\chapter{Sensors and instruments}
\label{vol01:ch05}

\epigraph{A sensor is a translator between the world and a number.
Its honesty is the engineer's first concern.}{The metrologist's
working rule, in the vocabulary of the VIM \cite{std:jcgm-vim} and
modern sensor handbooks \cite{text:fraden2016}.}

\chapmeta{Half-life: foundational. Archetypes: interface (sensor as
the calibration-style handoff between physical phenomenon and
indication, recurring from chapter 3), transport (the signal moving
from physical phenomenon through the transducer through the
conditioning circuit to the digitised quantity, first formal
introduction in this volume). Exercise target: 30. Project track:
physical with simulation alternative (hybrid). Hazard class: low.
Reader path default: standard, with sections explicitly tagged.
Named cases: Air France Flight 447 (2009), as a common-mode
sensor-failure case in the failure section; Ariane 5 Flight 501
(1996), as a common-mode software case in the chronic-class list;
Hubble Space Telescope primary mirror (1990), as the
instrument-testing-instrument case in the failure section;
Boeing 737 MAX MCAS (2018-2019) and Patriot missile system at
Dhahran (1991), at the short-form-summary level; Therac-25
(1985-1987), as the cross-domain analogue of shared-firmware
common-mode failure in the same-direction-failure subsection.}

\noindent
A measurement begins at a sensor. The sensor turns a physical
quantity into something the rest of the measurement chain can
operate on: a voltage, a current, a frequency, a digital count, a
visible deflection. A sensor is an engineering object: a physical
interface, a calibration function, a transport chain, and a failure
source. We judge it by the physical principle that produces the
signal, the specifications that bound the signal, the instrument
that carries it, and the shared causes that can make redundant
sensors fail together.

\Cref{fig:vol01:ch05:sensor-chain} sets the chapter's frame on a
single page: panel A shows the measurement chain from physical
phenomenon to digital record with the dominant noise or drift source
at each interface; panel B closes the loop through a reference
standard and the periodic recalibration that keeps the calibration
function alive; panel C shows the manufacturer's specification as
one term in a working uncertainty budget, alongside the reference
resistor, the ADC reference, quantisation, self-heating, and drift.
We return to each panel through the chapter; the figure also gives
the reader the shape of the sensor argument before any equation.

\input{volumes/01-quantity/ch05-sensors-and-instruments/figures/fig-sensor-chain.tex}

\section{What a sensor is}\pathtag{core}

A \engterm{sensor} is a device whose output is a known function of a
chosen physical quantity. A \engterm{transducer} is the more general
term for any device that converts one form of energy into another;
in working engineering vocabulary, a transducer that produces an
electrical (or, more generally, a measurable) output from a physical
input is a sensor. The two words are often used interchangeably; the
distinction matters in technical writing where a clear separation
between the energy-conversion mechanism (transducer) and its
metrological role (sensor) is needed.

Sensors do three jobs at once. They detect (the chosen physical
phenomenon is present and produces a signal), they discriminate (the
signal is distinguishable from noise and from other phenomena), and
they quantify (the signal magnitude maps to a numerical value of the
input quantity). A device that detects but does not quantify is a
\engterm{detector}; a device that quantifies but does not
discriminate from other phenomena is a poor sensor. Working sensors
do all three.

The output of a sensor is rarely the quantity of interest in its
final form. A thermistor's output is a resistance; the engineer
wants a temperature. A strain gauge's output is a resistance change;
the engineer first obtains strain, then obtains stress only after
applying a material model, geometry, and temperature compensation.
An accelerometer's output is a voltage; the engineer wants an
acceleration in metres per second squared. The mapping from sensor
output to quantity of interest is the \engterm{calibration function}
of the sensor, and it is what chapter 3 of this volume taught the
reader to acquire and document
\cite{std:jcgm-vim}.

\paragraph{Worked example: the household thermostat.} A wall
thermostat is a sensor by all three jobs at once. The bimetal coil
or the negative-temperature-coefficient (NTC) thermistor inside the
housing detects the air temperature (a thermal phenomenon is
present). The mechanical click of the bimetal at its set-point, or
the comparator threshold in the electronic unit, discriminates room
temperature from drafts on the wall behind the thermostat (signal
distinguishable from noise). The set-point dial or the digital
display maps the sensed temperature to a number the occupant can
read (signal to a numerical value). A thermostat that misses the
discriminate step (a unit mounted next to the kitchen oven that
treats the oven's heat as room heat) is a poor sensor for the room,
not because its physical principle is wrong, but because its
operational position fails to separate the signal it was bought to
measure from one it was not. Sensor selection sits in section~5.4;
selection failures of this kind are common.

\begin{archetype}[Interface, second appearance]
The interface archetype, introduced formally in chapter 3, recurs
here. A sensor is an interface between a physical quantity and a
representation that the rest of the measurement chain operates on.
The quality of the translation is the quality of the sensor; the
discipline of documenting the translation, with its uncertainty, is
the discipline of measurement. The sensor's interface is downstream
of nature and upstream of the calculation; the engineer who omits
the sensor's interface in their account of a measurement has not
explained where the number came from.
\end{archetype}

\begin{archetype}[Transport, first formal introduction]
A sensor's signal does not reach the engineer's calculation
directly. It travels from the transducer through a conditioning
circuit (amplification, filtering, isolation), often through an
analog-to-digital converter, often through a digital interface
(serial bus, network), and finally into the engineer's data record.
Each link in the transport chain is itself an interface; each can
add noise, drift, latency, or quantisation. The transport archetype
names the discipline of accounting for the entire chain, not only
the transducer. The transport archetype recurs throughout the work:
it appears in heat and mass transfer (Volume IV and Volume V), in
network packet routing (Volume VII), in system flows (Volume IX),
and in supply chains and infrastructure flows (Volume XII).
\end{archetype}

\section{Transducer principles: thermal, electrical, optical,
mechanical, chemical, magnetic}\pathtag{core}

Most working sensors translate quantities through six physical
families: thermal, electrical, optical, mechanical, chemical, and
magnetic. Fraden's handbook \cite{text:fraden2016} gives the longer
catalogue; the Webster--Eren measurement-systems handbook
\cite{text:webster2014} treats the spatial, mechanical, thermal, and
radiation sub-classes at depth; Bentley's measurement-systems text
\cite{text:ch05-bentley2005}, Doebelin and Manik
\cite{text:ch05-doebelin2007}, and Northrop
\cite{text:ch05-northrop2014} cover the conditioning electronics and
the static and dynamic specifications. This chapter keeps the
families a working engineer must recognise before reading a
datasheet. \Cref{fig:vol01:ch05:transducer-families} gives the
catalogue at a glance with two or three representative working
examples in each family;
\repopath{data/sensor_principles.csv} extends the list with measurand,
working range, typical accuracy, and readout column for each
example, in a form a reader can paste into a comparison spreadsheet
during selection (section~5.4).

\input{volumes/01-quantity/ch05-sensors-and-instruments/figures/fig-transducer-families.tex}

\subsection{Thermal}

Thermal sensors translate temperature into an electrical or
mechanical signal. The dominant working types span seven orders of
magnitude in cost and three orders of magnitude in accuracy.

\engterm{Thermocouples} produce a small voltage from the
junction of two dissimilar metals at a measured temperature
relative to a reference junction (the Seebeck effect). The
working types and their conditions are defined by IEC 60584-1:2013:

\begin{itemize}[leftmargin=*]
  \item \textit{Type J} (iron versus copper-nickel constantan):
    range $-40$ to $+750\,\si{\degreeCelsius}$; sensitivity
    $52\,\si{\micro\volt\per\kelvin}$ near room temperature;
    susceptible to oxidation of the iron leg above
    $\sim 500\,\si{\degreeCelsius}$.
  \item \textit{Type K} (chromel Ni-Cr versus alumel Ni-Al-Mn-Si):
    range $-200$ to $+1260\,\si{\degreeCelsius}$; sensitivity
    $41\,\si{\micro\volt\per\kelvin}$ near $25\,\si{\degreeCelsius}$,
    varying nonlinearly to $40\,\si{\micro\volt\per\kelvin}$ at
    $400\,\si{\degreeCelsius}$; the workhorse of industrial
    instrumentation.
  \item \textit{Type T} (copper versus copper-nickel constantan):
    range $-200$ to $+350\,\si{\degreeCelsius}$; sensitivity
    $43\,\si{\micro\volt\per\kelvin}$ at room temperature;
    preferred for sub-zero cryogenic service due to the linear
    response of copper at low temperatures.
  \item \textit{Type N} (Nicrosil versus Nisil): range $-200$ to
    $+1260\,\si{\degreeCelsius}$; sensitivity $26\,\si{\micro
    \volt\per\kelvin}$; improved stability over Type K in
    high-temperature long-duration service, the replacement
    standard in aerospace.
  \item \textit{Type R} (Pt-13Rh versus Pt) and \textit{Type S}
    (Pt-10Rh versus Pt): range $0$ to $+1768\,\si{\degreeCelsius}$;
    sensitivity $10\,\si{\micro\volt\per\kelvin}$; noble-metal
    types used for high-temperature reference work; Type S is the
    ITS-90 standard interpolation instrument from $630\,\si{
    \degreeCelsius}$ to the silver point.
  \item \textit{Type B} (Pt-30Rh versus Pt-6Rh): range $250$ to
    $+1820\,\si{\degreeCelsius}$; the highest-temperature working
    type, used in glass-forming and steel-melt instrumentation.
\end{itemize}
The NIST monograph 175 \cite{paper:ch05-nist-thermocouple-tables}
gives the working polynomial fits and the maximum operating
conditions for each type. Reference-junction compensation
is essential: a Type K reading at $300\,\si{\degreeCelsius}$ with
a $25\,\si{\degreeCelsius}$ reference junction generates $11.0\,\si{
\milli\volt}$ at the meter terminals, but the same measuring
junction at the same temperature with a $0\,\si{\degreeCelsius}$
reference (an ice-bath or an internal silicon thermometer
reading $0\,\si{\degreeCelsius}$ exactly) generates $12.21\,\si{
\milli\volt}$. The difference is the contribution of the
$25\,\si{\degreeCelsius}$ at the cold junction.

\engterm{Resistance temperature detectors (RTDs)} use the
temperature dependence of a metal's resistance. The standard is
platinum: Pt100 ($R_{0} = 100\,\si{\ohm}$ at $0\,\si{
\degreeCelsius}$) and Pt1000 ($R_{0} = 1000\,\si{\ohm}$) per
IEC 60751:2022. The Callendar-Van Dusen equation describes the
response: $R(T) = R_{0}[1 + A T + B T^{2} + C(T - 100)T^{3}]$ for
$T < 0\,\si{\degreeCelsius}$, with $A = 3.9083 \times 10^{-3}\,
\si{\per\kelvin}$, $B = -5.775 \times 10^{-7}\,\si{\per\kelvin
\squared}$, $C = -4.183 \times 10^{-12}\,\si{\per\kelvin\tothe{4}}$
(coefficient $C$ is zero for $T \geq 0$). Class B Pt100 tolerance
per IEC 60751 is $\pm (0.30 + 0.005|T|)\,\si{\kelvin}$ (so
$\pm 0.30\,\si{\kelvin}$ at $0\,\si{\degreeCelsius}$,
$\pm 0.80\,\si{\kelvin}$ at $100\,\si{\degreeCelsius}$), class A
$\pm (0.15 + 0.002|T|)\,\si{\kelvin}$, class AA $\pm (0.10 +
0.0017|T|)\,\si{\kelvin}$. Industrial 4-wire connections
eliminate lead-resistance error; 3-wire connections use a
bridge-compensation method that cancels lead resistance to
first order. Working range: $-200$ to $+850\,\si{\degreeCelsius}$.
\Cref{fig:vol01:ch05:rtd-leadwire} contrasts the three connection
schemes and the lead-resistance error each carries.

\input{volumes/01-quantity/ch05-sensors-and-instruments/figures/fig-rtd-leadwire.tex}

\paragraph{Worked example: RTD lead-resistance error.} A Pt100
class-B RTD sits at the end of a $20 \,\si{\meter}$ run of
$24$-AWG copper extension cable, one conductor out and one back.
Copper at $24$ AWG has resistance $\sim 84 \,\si{\milli\ohm\per
\meter}$, so each lead contributes $R_{L} = 20 \times 0.084 =
1.68 \,\si{\ohm}$ and the round trip is $2 R_{L} = 3.36 \,\si{
\ohm}$. Read as a 2-wire connection, the meter sees $R(T) + 2 R_{L}$
and attributes the extra resistance to temperature. Near
$0 \,\degC$ the Pt100 slope is $\dd R/\dd T = R_{0} \alpha =
100 \times 0.00385 = 0.385 \,\si{\ohm\per\kelvin}$, so
$3.36 \,\si{\ohm}$ maps to an apparent
$3.36/0.385 \approx 8.7 \,\degC$ of bias, dwarfing the class-B
tolerance of $\pm 0.30 \,\degC$ at $0 \,\degC$. The 3-wire
connection routes a third conductor of the same length and gauge;
a balanced bridge subtracts one $R_{L}$ from the measurement leg,
and the residual error is the \emph{mismatch} between the two
current-carrying leads, not the full lead resistance. For leads
matched to $\pm 2\,\%$ the residual is
$0.02 \times 1.68/0.385 \approx 0.09 \,\degC$, back inside the
class-B budget. The 4-wire connection forces the excitation current
through the outer pair and senses the voltage through the inner
pair, which carry only the voltmeter's input-bias current (of order
picoamperes); the lead drop vanishes to that current times the lead
resistance, roughly $10^{-12} \times 1.68 \approx 2 \,\si{\pico
\volt}$, negligible. The engineering lesson is that a Pt100 is a
$0.1 \,\degC$-class sensor only when its wiring is chosen to match
the budget: the same element on a long 2-wire run is a
several-degree instrument.

\engterm{Thermistors} use the strongly temperature-dependent
resistance of a doped metal-oxide semiconductor. NTC thermistors
(common in consumer applications) have characteristic
resistance $R_{0} = 10\,\si{\kilo\ohm}$ at $25\,\si{
\degreeCelsius}$ with $B$-parameter $B_{25/85} = 3950 \pm 5\%\,
\si{\kelvin}$ (a typical Vishay NTCLE100E3103JB0). The
$\beta$-parameter approximation is $R(T) = R_{0}\exp[B(1/T -
1/T_{0})]$ where $T$ and $T_{0}$ are in kelvin. The Steinhart-Hart
equation gives a working three-parameter fit:
$1/T = a + b\ln R + c(\ln R)^{3}$ with $a, b, c$ determined from
a three-point calibration (typically $0$, $25$, $100\,\si{
\degreeCelsius}$); the three-parameter fit reaches
$\pm 0.05\,\si{\kelvin}$ over a $0$-$100\,\si{\degreeCelsius}$
range. NTC range: typically $-50$ to $+150\,\si{\degreeCelsius}$.

\engterm{Infrared thermometers and pyrometers} use the thermal
radiation emitted by an object to infer temperature via Planck's
law. Working models: handheld single-pixel pyrometer (Fluke 568
at $\pm 1\,\si{\kelvin}$ in the $-40$ to $+800\,\si{
\degreeCelsius}$ range, $30$:$1$ optical resolution); two-colour
pyrometer (Optris CT 2MH1 at $\pm 0.5\%$ of reading, two
wavelengths to cancel emissivity effects); infrared imager
(FLIR T865 at $\pm 1\,\si{\kelvin}$ over $-40$ to $+1500\,\si{
\degreeCelsius}$ with $640 \times 480$-pixel microbolometer
detector). All require knowledge or measurement of the target's
emissivity; an emissivity error of $\pm 0.05$ from the assumed
value produces a reading error of $\sim \pm 7\,\si{\kelvin}$ at
$500\,\si{\degreeCelsius}$, comparable to the calibration
uncertainty.

The chapter project uses an NTC thermistor (a Vishay
NTCLE100E3103JB0 or equivalent), so the Steinhart-Hart and
$\beta$-parameter models will carry the calibration work.

\paragraph{Worked example: type-K cold junction.} A type-K
thermocouple with measuring junction at $T_{\text{m}}$ and reference
junction at $T_{\text{r}} = 0 \,\degC$ produces an open-circuit
voltage of order $S_{K}(T_{\text{m}} - T_{\text{r}})$ where
$S_{K} \approx 40 \,\si{\micro\volt\per\kelvin}$ near room
temperature \cite{text:fraden2016}. A measuring junction at
$T_{\text{m}} = 300 \,\degC$ produces approximately
$300 \times 40 = 12 \,\si{\milli\volt}$. A working type-K instrument
does not maintain a $0 \,\degC$ reference junction; it measures the
reference-junction temperature with a separate sensor (an integrated
silicon thermometer in modern handheld thermocouple readers) and
compensates digitally. Cold-junction-compensation error is therefore
itself a thermometer error, and a handheld thermocouple reader's
quoted accuracy is the combination of its thermocouple amplifier and
its internal cold-junction sensor. A reader who omits the
cold-junction sensor (a thermocouple plugged directly into a
millivoltmeter) has the reading offset by the room temperature; a
$25 \,\degC$ room offsets the $300 \,\degC$ reading by
$25 \times 40 = 1.0 \,\si{\milli\volt}$, which the linear inverse
reports as a $25 \,\degC$ measurement bias.

\subsection{Electrical}

Electrical sensors translate electrical quantities into signals
suitable for further measurement. The dominant types span DC to
RF in operating frequency and span six orders of magnitude in
current range.

\engterm{Voltage sensors}. The standard voltage transducer is the
resistive divider, with the upper resistor matched to a precision
ADC reference. Bulk-foil resistors (Vishay Z201) reach
$\pm 0.005\%$ tolerance with $0.05\,\si{ppm/\kelvin}$ temperature
coefficient; thin-film networks (Vishay PNM Series) reach matched
ratio stability of $0.5\,\si{ppm/\kelvin}$. For voltages above the
ADC's input range, capacitor-coupled dividers handle AC while
high-voltage probes (Tektronix P5100A at $1000:1$, $250\,\si{
\mega\hertz}$ bandwidth, $\$1200$) handle short-duration
transients. The discipline at every level is to size the divider
so the input impedance does not load the source significantly: a
$10\,\si{\mega\ohm}$ divider on a $1\,\si{\kilo\ohm}$ source
loads the source by $0.01\%$, acceptable for most working
instruments.

\engterm{Current sensors}. Four working principles cover the
practical range. \engterm{Shunt resistors} place a precision
low-value resistor in series with the load; the voltage across
the shunt scales with current per Ohm's law. Working values:
$10\,\si{\milli\ohm}$ for $\pm 100\,\si{\ampere}$ at $\pm 1\,\si{
\volt}$ output, $100\,\si{\micro\ohm}$ for kiloampere service
(Vishay WSR-2 manganin shunts, $\pm 0.1\%$ initial tolerance,
$\pm 75\,\si{ppm/\kelvin}$). \engterm{Current transformers (CTs)}
use magnetic induction in a toroidal core; the primary is the
load conductor passed through the core, the secondary is many
turns of fine wire feeding a burden resistor. Ratios from $5:5$
to $5000:5$, accuracy classes $0.1$, $0.2$, $0.5$, $1.0$ per
IEC 61869-2:2012. \engterm{Hall-effect current sensors} measure
the magnetic field of the conductor with a Hall element in a slot
of a flux-concentrating core; the LEM HAS 50-S at $\pm 50\,\si{
\ampere}$ range with $\pm 1\%$ accuracy and $\pm 0.05\,\si{
\milli\tesla}$ offset is the working DC capable industrial unit.
\engterm{Rogowski coils} are flexible air-core windings that
measure $\dd I/\dd t$ and integrate to current; PEM CWT-15 at
$300\,\si{\ampere}$ peak, $1\,\si{\mega\hertz}$ bandwidth,
\$800.

\engterm{Resistance sensors}. Four-wire (Kelvin) connection is
the working architecture: the current-source leads carry the
test current, the voltage-sense leads measure the voltage drop
across the device under test without lead-resistance error. Bench
DMMs such as the Keysight 34470A at 7.5-digit resolution measure
resistance from $1\,\si{\micro\ohm}$ (with four-wire and
$10\,\si{\ampere}$ test current) through $100\,\si{\mega\ohm}$
with two-wire connection. Above $100\,\si{\mega\ohm}$, megohm
meters (Megger MIT525 at $\$5000$, $50\,\si{\volt}$ to $5\,\si{
\kilo\volt}$ programmable test voltage) measure insulation
resistance through $10\,\si{\tera\ohm}$ with the test voltage
selected to expose the dielectric's voltage-dependent breakdown.

\engterm{Charge sensors}. The classical electrometer measures
charge by integrating a low-current input onto a known
capacitance. Modern instruments such as the Keysight B2987B
(femtoampere noise, $10^{-15}\,\si{\coulomb}$ resolution, \$$20{,}000$)
measure low-current photodiode currents, ion-chamber outputs,
piezoelectric charge sensors, and electrostatic charge on
isolated conductors. The working principle is the high-impedance
charge amplifier: an op-amp with capacitive feedback $C_{f}$
gives $V_{\text{out}} = -Q/C_{f}$; with $C_{f} = 100\,\si{\pico
\farad}$, a $10\,\si{\pico\coulomb}$ input produces a $100\,\si{
\milli\volt}$ output, comfortable for a downstream ADC. Below
$10\,\si{\femto\coulomb}$, the working uncertainty is set by the
amplifier's voltage and current noise referred to the input
through the feedback impedance.

\engterm{Capacitance sensors}. The Lion Precision CPL290 at
$\pm 100\,\si{\micro\meter}$ range with $4\,\si{\nano\meter}$
resolution is the working industrial sub-micrometre displacement
sensor; the Micro-Epsilon capaNCDT 6500 covers the same
displacement metrology at $\$5000$. Capacitive level sensors
(Endress+Hauser Liquicap M FMI51 at $\pm 5\,\si{\milli\meter}$ on
$0$-$5\,\si{\meter}$ levels) detect liquid level by the
dielectric-constant difference between the liquid and the gas
above it; the working uncertainty is set by liquid composition
variability rather than by the capacitance measurement itself.
Capacitive touch sensors in consumer electronics (Cypress
PSoC 4 family) detect finger contact at $\sim 0.5\,\si{\pico
\farad}$ change against an empty-electrode baseline of $\sim 5$-$
20\,\si{\pico\farad}$.

A working principle of electrical-quantity sensing is that the
sensor must not significantly perturb the circuit it is measuring.
A voltmeter with low input impedance loads the circuit; a current
sensor with high series impedance changes the current it is trying
to measure. The discipline of \engterm{loading effects} is taken
up at depth in Volume IV's electrical-circuits chapters.

\paragraph{Worked example: shunt vs current transformer.} A
$10\,\si{\milli\ohm}$ precision shunt reading $100\,\si{\ampere}$
develops $1\,\si{\volt}$ across the shunt, dissipates $100\,\si{
\watt}$, and inserts a $10\,\si{\milli\ohm}$ series impedance into
the circuit it measures; the working uncertainty is set by the
shunt's thermal coefficient ($\pm 75\,\si{ppm/\kelvin}$ for the
WSR-2; at $30\,\si{\degreeCelsius}$ self-heating above ambient,
$\Delta R/R = 75 \times 10^{-6} \times 30 = 0.23\%$) and its
self-heating. A current transformer reading the same $100\,\si{
\ampere}$ at a $100:1$ ratio produces a secondary current of
$1\,\si{\ampere}$ through a burden resistor, dissipates negligible
primary-side power, isolates the readout from the primary
circuit through the transformer's galvanic isolation, but works
only for alternating currents above its low-frequency cutoff
(typically $5\,\si{\hertz}$ for a $50\,\si{\hertz}$ CT design).
The engineering trade-off, conduction loss and DC capability
against isolation and AC-only operation, drives the selection;
either device is a sensor for current with the same datasheet
vocabulary. For DC capability plus isolation, the working choice
is the Hall-effect or flux-gate current sensor; for AC at low cost
with no DC, the CT; for DC at lowest cost, the shunt.

\subsection{Optical}

Optical sensors translate light (or other electromagnetic
radiation) into electrical signals. The dominant types span ten
orders of magnitude in light level and four decades in cost.

\engterm{Photodiodes} convert incident photons into charge
carriers via the photoelectric effect in a semiconductor junction.
Material and bandgap select the spectral range: silicon
($E_{g} = 1.12\,\si{\electronvolt}$, cutoff $\lambda \approx
1.1\,\si{\micro\meter}$) covers $200$-$1100\,\si{\nano\meter}$ with
responsivity peaking at $0.55$-$0.65\,\si{\ampere\per\watt}$
between $700$ and $900\,\si{\nano\meter}$ (Hamamatsu S1226,
Hamamatsu S1227 series); InGaAs ($E_{g} = 0.74\,\si{\electronvolt}$)
covers $900$-$1700\,\si{\nano\meter}$ at $0.9\,\si{\ampere\per
\watt}$ peak (Hamamatsu G12180); germanium ($E_{g} = 0.66\,\si{
\electronvolt}$) covers $800$-$1800\,\si{\nano\meter}$ but with
worse noise than InGaAs; extended-InGaAs reaches $2.6\,\si{
\micro\meter}$. Avalanche photodiodes (APDs) provide internal
gain $M = 10$-$200$ through impact ionisation but with
excess-noise factor $F = M^{x}$ for $x \approx 0.3$-$0.5$
depending on material; the working choice for shot-noise-limited
detection at low light. Photodiodes are fast (rise time
$1\,\si{\nano\second}$ for $1\,\si{\milli\meter}^{2}$ silicon
PIN), linear over six decades in light level (Thorlabs FDS100 at
$10\,\si{\pico\watt}$ to $10\,\si{\milli\watt}$), and cheap (\$10
to \$200).

\engterm{Phototransistors} integrate a bipolar transistor with a
photodiode at its base junction, giving current gain $\beta = 100$
to $1000$ in exchange for slower response time ($\sim 10\,\si{
\micro\second}$ rise) and nonlinear collector-current versus
optical-power transfer. The working choice when current gain
matters and bandwidth does not (consumer-electronics ambient-
light sensing).

\engterm{Photomultiplier tubes} amplify weak optical signals
through cascaded secondary-electron emission at $8$-$14$ dynode
stages. Photocathode quantum efficiency at peak wavelength is
$20$-$25\%$ for bialkali (Hamamatsu R6095 in the $300$-$650\,\si{
\nano\meter}$ band) and $30$-$40\%$ for GaAsP (Hamamatsu H11526 in
the same range). Total gain $G = 10^{5}$ to $10^{7}$ at supply
voltages $700$-$2500\,\si{\volt}$ depending on the dynode chain
voltage divider. Dark-current contributions: thermionic emission
from the photocathode ($\sim 100\,\text{counts/s}$ at room
temperature for bialkali, falling to $\sim 1\,\text{count/s}$ at
$-30\,\si{\degreeCelsius}$), Cherenkov radiation in the glass
envelope from cosmic muons, and afterpulses from ion bombardment.
Modern photon counting modules (Hamamatsu C9744, ID Quantique
ID120) integrate the tube with high-voltage supply and counter
electronics for plug-and-play operation. PMT working
applications: scintillation gamma-ray spectroscopy, biomedical
luminescence, low-light astronomy, single-photon-detection
quantum optics. Trade-offs against silicon photomultipliers
(SiPMs, Hamamatsu MPPC): SiPMs are smaller, magnetic-field-
tolerant, and use $30$-$70\,\si{\volt}$ bias; PMTs offer lower
dark count rates and larger detection area.

\engterm{Image sensors} (CCD, CMOS) are two-dimensional arrays of
photodetectors. CCDs (Charge-Coupled Devices) shift photo-charge
through the array to a single output amplifier, achieving high
charge-transfer efficiency ($> 99.999\%$ per shift) and uniform
response; the read noise floor of a scientific cooled CCD is
$2$-$5\,e^{-}$ rms (Andor iKon-M at $-100\,\si{\degreeCelsius}$
two-stage thermoelectric cooling, $0.001\,e^{-}/\text{pixel}/
\text{s}$ dark current). CMOS sensors integrate the readout
amplifier into each pixel, enabling higher frame rates ($30$-
$1000\,\si{fps}$ for industrial machine vision; $10000\,\si{fps}$
for ultra-high-speed scientific) and global-shutter operation,
at the cost of pixel-to-pixel non-uniformity and higher read
noise ($1$-$10\,e^{-}$ rms for back-illuminated sCMOS such as the
Hamamatsu ORCA-Flash4.0). Smartphones use rolling-shutter CMOS
with $40$-$50\,\si{\milli\meter}^{2}$ sensor area divided into
$48$-$200$ megapixels (Sony IMX989, Samsung GN5); the read noise
is $3$-$5\,e^{-}$ rms and dynamic range is $\sim 60\,\si{
\decibel}$.

\engterm{Fibre-optic sensors} embed sensing elements in optical
fibres. Fibre Bragg grating (FBG) sensors detect strain or
temperature through the wavelength shift $\Delta\lambda/\lambda =
(1 - p_{e})\varepsilon + (\alpha + \zeta)\Delta T$ where
$p_{e} \approx 0.22$ is the photo-elastic coefficient for silica,
$\alpha \approx 0.55\,\si{ppm/\kelvin}$ the thermal expansion,
and $\zeta \approx 8.6\,\si{ppm/\kelvin}$ the thermo-optic
coefficient. A standard FBG (Smart Fibres SmartScan SBI-Plus, $4$
or $16$ channels at $2.5\,\si{\kilo\hertz}$ scan rate) reaches
$1\,\si{\micro\strain}$ strain resolution or $0.1\,\si{\kelvin}$
temperature resolution. Distributed-acoustic-sensing (DAS,
Silixa iDAS Plus) and distributed-temperature-sensing (DTS,
Sensornet Halo) systems use Brillouin or Raman backscatter to
profile strain or temperature continuously along $50\,\si{\kilo
\meter}$ of fibre at $1\,\si{\meter}$ spatial resolution.

\engterm{Laser interferometers} measure displacement and velocity
by interference between a reference beam and a beam reflected
from the target. Working instruments: HP/Keysight 5530 with
helium-neon laser at $632.991\,2\,\si{\nano\meter}$, $0.5\,\si{
\nano\meter}$ resolution, sub-ppm displacement accuracy; Renishaw
XL-80 at the same wavelength and accuracy class, with rotary,
straightness, and squareness measurement options. These are the
primary metrology instruments for length at the sub-micrometre
scale (chapter 3 of this volume).

\paragraph{Worked example: smartphone proximity sensor.} The
smartphone proximity sensor, the device that turns the screen off
during a phone call when the phone is held to the user's ear, is a
short-range infrared system: a low-power IR LED illuminates the
nearby surface, a silicon photodiode detects the reflected light,
and a comparator threshold reports near / far. The sensor measures
proximity by inference from reflected optical power; the calibration
function depends on the reflectivity of the user's face, which the
device does not know. The working solution is a threshold set well
below the optical power of a typical face return, with hysteresis to
prevent flutter; the resulting device is a discriminator, not a
quantifier (proximity is reported as near or far, not as a distance
in millimetres). The sensor reports a fact (the face is present) by
measuring a quantity (reflected optical power) and discarding the
quantitative information. A reader who tried to use the same sensor
to measure a distance would have built the wrong instrument for the
job.

\paragraph{Worked engineering case: photodiode shot-noise
analysis for a low-light optical receiver.} A common transducer
problem reads: select and condition a silicon photodiode for a
near-infrared optical receiver whose input optical power is
$P_{\text{opt}} = 1 \,\si{\micro\watt}$ and whose required
electrical signal-to-noise ratio at the post-amplifier output is
$30 \,\si{\decibel}$ over a $1 \,\si{\mega\hertz}$ bandwidth. The
worked solution moves through the four noise contributions in
order and identifies the regime in which each dominates.

\textbf{Step 1: convert optical power to photocurrent.} A silicon
photodiode at $\lambda = 850 \,\si{\nano\meter}$ has responsivity
$R_{\lambda} \approx 0.55 \,\si{\ampere\per\watt}$, set by the
quantum efficiency near $80\,\%$ and the photon energy $h c/
\lambda$. The photocurrent is
$I_{\text{ph}} = R_{\lambda} P_{\text{opt}} = 0.55 \times 10^{-6}
= 0.55 \,\si{\micro\ampere}$.

\textbf{Step 2: compute the shot-noise current.} Shot noise on
the photocurrent has spectral density
$i_{\text{shot}}^{2}/\Delta f = 2 e I_{\text{ph}}$, with $e =
1.602 \times 10^{-19} \,\si{\coulomb}$ the electron charge. The
integrated rms over the $1 \,\si{\mega\hertz}$ bandwidth is
$i_{\text{shot}} = \sqrt{2 e I_{\text{ph}} \Delta f} =
\sqrt{2 \times 1.602 \times 10^{-19} \times 5.5 \times 10^{-7}
\times 10^{6}} = \sqrt{1.76 \times 10^{-19}} \approx 4.2 \times
10^{-10} \,\si{\ampere} = 0.42 \,\si{\nano\ampere}$.

\textbf{Step 3: compare against dark-current shot noise.} A
representative small-area silicon photodiode has dark current
$I_{d} \approx 1 \,\si{\nano\ampere}$ at room temperature
\cite{text:ch05-pallas-areny-webster2001}. Dark-current shot noise
contributes $i_{\text{dark}} = \sqrt{2 e I_{d} \Delta f} =
\sqrt{2 \times 1.602 \times 10^{-19} \times 10^{-9} \times 10^{6}}
= \sqrt{3.2 \times 10^{-22}} \approx 1.8 \times 10^{-11} \,\si
{\ampere} = 18 \,\si{\pico\ampere}$, more than twenty times
smaller than the signal shot noise. The dark current contributes
to the noise floor only when the optical signal drops by roughly
$(I_{d}/I_{\text{ph}})^{2} = (1.0/0.55)^{2} \approx 3.3$ below
the present operating point.

\textbf{Step 4: compute the transimpedance-amplifier noise.} A
working transimpedance amplifier (TIA) uses a feedback resistor
$R_{f}$ and an op-amp with input-referred current noise $i_{n}$
and voltage noise $e_{n}$. Choose $R_{f}$ so that the photocurrent
sits at a useful fraction of the op-amp's output range: for an
output target $V_{\text{out}} = 1 \,\si{\volt}$ at $I_{\text{ph}}
= 0.55 \,\si{\micro\ampere}$, $R_{f} = V_{\text{out}}/I_{\text{ph}}
\approx 1.8 \,\si{\mega\ohm}$. The Johnson-noise current of
$R_{f}$ is $i_{R_{f}} = \sqrt{4 k_{B} T \Delta f / R_{f}} =
\sqrt{4 \times 1.38 \times 10^{-23} \times 295 \times 10^{6}/(1.8
\times 10^{6})} \approx 9.5 \times 10^{-11} \,\si{\ampere} = 95
\,\si{\pico\ampere}$. A modern low-noise CMOS op-amp contributes
$i_{n} \approx 1 \,\si{\femto\ampere}/\sqrt{\si{\hertz}}$ and
$e_{n} \approx 5 \,\si{\nano\volt}/\sqrt{\si{\hertz}}$; over the
$1 \,\si{\mega\hertz}$ bandwidth, $i_{n} \sqrt{\Delta f} = 1 \,\si
{\pico\ampere}$ and $e_{n}/R_{f} \times \sqrt{\Delta f} =
5/1.8 \times 10^{-6} \times \sqrt{10^{6}} = 2.8 \,\si{\nano\ampere}$.
The voltage noise dominates because the feedback resistor is
large; the engineer who reduces $R_{f}$ to lower the voltage-noise
contribution must accept a smaller output swing or add a second
gain stage.

\textbf{Step 5: combine the noise components.} The four
contributions (signal shot, dark shot, Johnson of $R_{f}$, op-amp
voltage noise referred to current) combine in quadrature:
\[
i_{\text{noise,total}} = \sqrt{i_{\text{shot}}^{2} +
i_{\text{dark}}^{2} + i_{R_{f}}^{2} + (e_{n} \sqrt{\Delta f}/R_{f})^{2}}.
\]
Substituting, $i_{\text{noise,total}} = \sqrt{(0.42)^{2} +
(0.018)^{2} + (0.095)^{2} + (2.8)^{2}} \,\si{\nano\ampere}
\approx 2.8 \,\si{\nano\ampere}$. The op-amp voltage noise
referred through $R_{f}$ dominates the total by a factor of
seven over the signal shot noise.

\textbf{Step 6: compute SNR and meet the specification.} The
signal-to-noise ratio in current is $\mathrm{SNR} =
I_{\text{ph}}/i_{\text{noise,total}} = 0.55 \,\si{\micro\ampere}/
2.8 \,\si{\nano\ampere} \approx 196$, or $20 \log_{10}(196)
\approx 45.9 \,\si{\decibel}$, $15.9 \,\si{\decibel}$ above the
$30 \,\si{\decibel}$ requirement. The design has comfortable
margin; the lowest-cost optimisation is to lower the bandwidth
toward the actual signal content, since shot and voltage noise
both scale as $\sqrt{\Delta f}$ and dropping the bandwidth to
$100 \,\si{\kilo\hertz}$ would gain $10 \,\si{\decibel}$ of SNR
at the cost of slower response.

\textbf{Sanity check.} At the shot-noise limit (assuming
perfectly noiseless amplifier and zero dark current), the SNR
would be $I_{\text{ph}}/i_{\text{shot}} = 0.55 \,\si{\micro
\ampere}/0.42 \,\si{\nano\ampere} \approx 1310$, or $62 \,\si
{\decibel}$. The amplifier-dominated design sits $16 \,\si
{\decibel}$ below this fundamental floor. The improvement path is
either a lower-noise op-amp (cutting $e_{n}$ by a factor of two
buys $3 \,\si{\decibel}$ of SNR), an avalanche photodiode (gains
$10$ to $20 \,\si{\decibel}$ at the cost of voltage and
temperature stability), or a photomultiplier (for the lowest-
light regime where shot noise itself is dominant). For the
project's $30 \,\si{\decibel}$ target the simple silicon-photodiode
TIA is adequate; for a $50 \,\si{\decibel}$ target an APD would
be required \cite{text:ch05-pallas-areny-webster2001}.

\textbf{Common error.} Quoting SNR at the shot-noise limit without
accounting for amplifier noise. The fundamental limit is rarely
reached in practice; the working SNR is set by whichever
contribution dominates the quadrature sum. The engineer who
designs to the shot-noise limit and discovers $15 \,\si{\decibel}$
of missing margin in the lab has made exactly this mistake. The
working discipline is to compute all four contributions, identify
the dominant term, and design the next iteration around reducing
it. The same discipline applies to thermal noise in mechanical
sensors and to phase noise in oscillators (chapter 6 of this
volume); the structural pattern transfers across domains.

\subsection{Mechanical}

Mechanical sensors translate mechanical quantities (force,
pressure, displacement, acceleration, angular rate) into
measurable signals. The dominant types span four to six orders of
magnitude in cost.

\engterm{Strain gauges} are foil or semiconductor elements
bonded to a structure; their resistance changes with applied
strain through the gauge-factor relation $\Delta R/R = G\,
\varepsilon$. Foil gauges (constantan or karma alloy on a
polyimide carrier) have $G = 2.0 \pm 1\%$ at $R_{0} = 120$, $350$,
$700$, or $1000\,\si{\ohm}$ nominal resistance per the Vishay
Micro-Measurements catalogue; the workhorse $R_{0} = 350\,\si{
\ohm}$ EA-13-125AC-350 reaches strain measurements from
$\sim 10^{-6}$ to $5 \times 10^{-3}$ over its $-75$ to
$+200\,\si{\degreeCelsius}$ working range. Semiconductor (silicon)
strain gauges have $G = 100$ to $200$ but with strong nonlinearity
and temperature sensitivity that limit their working range.
Bridge architectures (quarter, half, full Wheatstone) transform
the resistance change into a voltage; the full-bridge form with
two active legs in tension and two in compression delivers four
times the single-arm output and cancels first-order thermal
effects \cite{text:ch05-pallas-areny-webster2001}.

\engterm{Piezoelectric sensors} produce charge under applied
stress through the direct piezoelectric effect $Q = d_{ij}\,F$,
where $d_{ij}$ is the piezoelectric charge constant. Quartz
($d_{11} = 2.3\,\si{\pico\coulomb\per\newton}$) is the working
material for precision metrology because its piezoelectric
response is temperature-stable to $\pm 0.01\%/\si{\kelvin}$ in
its working range. Lead-zirconate-titanate (PZT) ceramics offer
much higher sensitivity ($d_{33} = 200$ to $600\,\si{\pico
\coulomb\per\newton}$ depending on composition) but with
temperature drift around $\pm 2\%/\si{\kelvin}$ near Curie
temperatures. Polyvinylidene fluoride (PVDF) is flexible,
inexpensive, and tolerates large strains; $d_{31} = 23\,\si{
\pico\coulomb\per\newton}$. Working instruments: PCB Piezotronics
353B Series quartz shear accelerometers ($100\,\si{\milli\volt
\per g}$ sensitivity, $0.5$ to $5000\,\si{\hertz}$ bandwidth,
$\pm 0.5\,g$ resolution at $1\,\si{\hertz}$, $\$800$); Kistler
8703B vibrating-coil quartz triaxial accelerometers ($\$2500$).
With charge-mode (electrometer) readout, the high input impedance
gives a low-frequency cutoff at $1/(2\pi R_{f} C_{f})$ where
$R_{f}$ and $C_{f}$ are the charge amplifier's feedback
components; for $R_{f} = 10^{12}\,\si{\ohm}$, $C_{f} = 100\,\si{
\pico\farad}$ the cutoff is $1.6\,\si{\milli\hertz}$, making
quasi-static measurements possible if the amplifier is
correspondingly designed. Voltage-mode (IEPE: integrated-
electronics-piezoelectric) readout uses an internal FET buffer
powered by a constant-current supply through the same cable; the
sensitivity is fixed, the low-frequency cutoff is around
$0.5\,\si{\hertz}$, but the system is rugged and cable-tolerant.

\engterm{Linear variable differential transformers (LVDTs)}
produce an AC voltage proportional to the displacement of a
ferromagnetic core within a primary-and-two-secondary-coil
assembly excited at $1$-$10\,\si{\kilo\hertz}$. With no sliding
electrical contact at the moving core, they tolerate dirty
industrial environments. Stroke ranges from $\pm 0.5\,\si{\milli
\meter}$ to $\pm 500\,\si{\milli\meter}$; non-linearity over the
specified range is $\pm 0.25\%$ of full scale for industrial
units (Macro Sensors HSAR series) and $\pm 0.05\%$ for precision
units (Schaevitz HSAR-750). Working sensitivity: $200\,\si{\milli
\volt\per\milli\meter}$ at $\pm 25\,\si{\milli\meter}$ stroke.
Industrial application: turbine-blade tip clearance, hydraulic
piston position, machine-tool tool position.

\engterm{MEMS accelerometers and gyroscopes} are micro-machined
inertial sensors using capacitive or piezoresistive readout of a
proof mass or vibrating structure. Consumer-grade MEMS such as
the STMicroelectronics LSM6DSO ($\$3$, $\pm 16\,g$ range,
$\pm 4000\,\si{\degree\per\second}$ gyro, $14$-bit output) deliver
$70\,\si{\micro g/\sqrt{\si{\hertz}}}$ noise density on the
accelerometer and $5\,\si{\milli\degree\per\second\per\sqrt{
\hertz}}$ on the gyro. Industrial MEMS such as the Analog Devices
ADXL355 ($\$30$, $\pm 8\,g$ range, $25\,\si{\micro g/\sqrt{\hertz}}$
noise) reach navigation-grade performance for guided-missile
inertial measurement units (IMUs) when combined with GPS aiding.
Strategic-grade ring laser gyros and fibre-optic gyros (Honeywell
GG1320 at $0.001\,\si{\degree\per\hour}$ bias stability,
$\$50{,}000$) operate at the boundary of what inertial-only
navigation can deliver. We take up MEMS as a discipline in
Volume VIII chapter 12.

\engterm{Pressure sensors} use a diaphragm whose deflection is
read by a strain gauge, capacitance, or piezoelectric element.
Absolute (referenced to vacuum), gauge (referenced to local
atmospheric), differential (between two ports), and sealed-gauge
(referenced to a sealed reference cavity, typically near
$1\,\text{atm}$) configurations cover all working applications.
Specific instruments: Honeywell 26PCFFB6T at $\pm 0.25\%$ FS
accuracy over $0$-$100\,\text{psi}$ gauge ($\$80$); Druck
PMP1400 industrial-grade with $\pm 0.04\%$ FS over $0$ to
$3.5\,\si{\mega\pascal}$ gauge ($\$800$); Heise PTE-1 calibration-
grade with $\pm 0.01\%$ FS at the same range ($\$5000$); Mensor
CPC8000 deadweight tester calibration source at $\pm 0.005\%$ FS
($\$50{,}000$). The chapter's working-laboratory pressure
instrument is the Heise or equivalent, calibrated annually
against the deadweight tester held at the local accredited
calibration laboratory.

\paragraph{Worked example: strain-gauge load cell.} A
foil strain gauge with gauge factor $G = 2$, unstrained resistance
$R = 350 \,\si{\ohm}$, bonded to a steel beam under axial strain
$\varepsilon = 10^{-3}$ (one millistrain, the working full-scale
strain for a typical load cell), produces a resistance change
$\Delta R = G R \varepsilon = 2 \times 350 \times 10^{-3} =
0.70 \,\si{\ohm}$, a fractional change of $0.2\,\%$. Read through a
Wheatstone bridge excited at $5 \,\si{\volt}$ in a single-active-arm
configuration, the bridge output is approximately
$V_{\text{br}} \approx V_{\text{ex}} \Delta R / (4 R) = 5 \times 0.7
/ (4 \times 350) \approx 2.5 \,\si{\milli\volt}$. The output is
about half the dynamic range of a $\pm 5 \,\si{\milli\volt}$
instrumentation amplifier and well above the Johnson-noise floor of
the bridge resistors over a $10 \,\si{\hertz}$ bandwidth. A four-arm
bridge with two active gauges in tension and two in compression
quadruples the output, eliminates the bridge's thermal common-mode
response, and is the working production configuration for load cells
and pressure transducers \cite{text:ch05-pallas-areny-webster2001}.

\subsection{Chemical}

Chemical sensors translate the presence or concentration of a
chemical species into an electrical signal. The dominant types
span the working analytical chemistry of clinical, environmental,
industrial, and consumer measurement.

\engterm{Ion-selective electrodes (ISEs)} produce a voltage whose
magnitude depends logarithmically on the activity of a target
ion through the Nernst equation $E = E_{0} + (RT/zF)\ln(a)$,
where $z$ is the ion's charge, $F = 96\,485\,\si{\coulomb\per
\mol}$ is the Faraday constant, $R = 8.314\,\si{\joule\per\mol
\per\kelvin}$ is the gas constant, and $a$ is the ion's activity
(concentration corrected for ionic strength). At $T = 25\,\si{
\degreeCelsius}$, the Nernstian slope is $(RT/F)\ln 10 \approx
59.16\,\si{\milli\volt}$ per decade for monovalent ions and
$29.58\,\si{\milli\volt}$ per decade for divalent. The pH
electrode (combined glass and silver/silver-chloride reference,
Mettler Toledo InLab Solids Pro-ISM at \$$250$ or laboratory-
grade Beckman Coulter PHI 5300 at \$$1500$) measures pH over $0$
to $14$ at $\pm 0.01$ accuracy after two-point calibration
against NIST-traceable buffers. Specific ISEs: Cole-Parmer
$\mathrm{Na}^{+}$ ISE at $\pm 0.05$ pNa, $\mathrm{F}^{-}$ ISE for
water-fluoridation monitoring at $\pm 1\,\si{\milli\gram\per
\liter}$ (Thermo Orion 9609BNWP), $\mathrm{NO}_{3}^{-}$ ISE for
agricultural soil analysis, $\mathrm{Ca}^{2+}$ ISE for clinical
chemistry. The standard interference ion (e.g., $\mathrm{K}^{+}$
interfering with the $\mathrm{Na}^{+}$ ISE through a selectivity
coefficient $K_{\text{Na,K}}^{\text{sel}} \approx 10^{-3}$)
requires application-specific calibration.

\engterm{Amperometric electrochemical sensors} read a current
proportional to analyte concentration at a poised working
electrode versus a reference. The Clark dissolved-oxygen electrode
(YSI Pro20i at \$$2000$) holds a platinum cathode at $-0.7\,\si{
\volt}$ versus a silver chloride reference; oxygen diffuses
through a Teflon membrane and reduces at the cathode, producing
a current proportional to dissolved oxygen ($i = 4FAD C/\delta$
where $A$ is the cathode area, $D$ the diffusivity in the
membrane, $C$ the dissolved-oxygen concentration, and $\delta$ the
membrane thickness). Glucose monitors (Abbott FreeStyle Libre 3
continuous glucose monitor at \$$70$ per sensor, $14$-day wear)
use glucose oxidase to convert glucose to gluconolactone plus
hydrogen peroxide; the peroxide is oxidised at the working
electrode producing a current proportional to glucose
concentration. Working accuracy: MARD (mean absolute relative
difference vs reference YSI laboratory glucose analyser)
$\sim 8\%$ for current-generation continuous glucose monitors.

\engterm{Optical chemical sensors} read absorbance, fluorescence,
or surface-plasmon-resonance (SPR) changes induced by chemical
binding events. UV-visible spectrophotometers (Thermo Scientific
Genesys 50 at \$$3000$ for $190$-$1100\,\si{\nano\meter}$) read
absorbance to $\pm 0.003$ A units; fluorescence spectrometers
(Horiba FluoroMax at \$$50{,}000$) detect single-molecule
quantities of fluorophores. SPR instruments (Cytiva Biacore 8K at
\$$300{,}000$) measure binding-induced refractive-index changes
at gold-film-functionalised surfaces; the detection limit reaches
$\sim 10\,\si{\pico\gram\per\milli\meter\squared}$ surface
coverage for the biopharmaceutical-affinity working application.
Optical chemical sensors share the optical-readout chain with
the optical sensors of the previous section but with the chemistry
of the binding event determining selectivity.

\engterm{Solid-state gas sensors}. Metal-oxide semiconductor
sensors (Figaro TGS family for $\mathrm{CO}$, $\mathrm{NO}_{x}$,
methane, hydrogen) use the reversible adsorption of reducing or
oxidising gases on a heated tin-dioxide ($\mathrm{SnO}_{2}$)
surface; adsorption changes the carrier concentration and hence
the bulk resistance. The Sensirion SGP41 multi-gas indoor air
quality sensor (\$$15$) measures total VOCs and $\mathrm{NO}_{x}$
through twin metal-oxide sensors at $\sim 1$ ppb resolution.
Electrochemical gas cells (Alphasense CO-B4 carbon monoxide
sensor at \$$60$, $0$-$500$ ppm, $\pm 5\%$ accuracy) use a
working-counter-reference electrode chemistry similar to the
Clark electrode but in air; the diffusion-limited reaction at the
working electrode produces a current proportional to gas
concentration. Photo-ionisation detectors (PIDs, RAE Systems
ppbRAE 3000+ at \$$5000$) ionise low-ionisation-potential VOCs
with a UV lamp ($10.6\,\si{\electronvolt}$ krypton lamp) and
measure the resulting ion current. Non-dispersive infrared (NDIR)
cells (Sensirion SCD41 for $\mathrm{CO}_{2}$ at \$$40$,
$0$-$40\,000$ ppm, $\pm 50$ ppm or $\pm 5\%$ accuracy) read the
infrared absorption at the analyte's molecular vibration band.

\paragraph{Worked example: two-point span-gas calibration of an
NDIR sensor.} An NDIR $\mathrm{CO}_{2}$ sensor reads infrared
absorption at the $4.26 \,\si{\micro\meter}$ band and inverts the
Beer--Lambert law to recover concentration. In practice the
manufacturer linearises the response so that the reported reading
$y$ relates to the true concentration $x$ by a working affine model
$y = m x + b$, with $m$ the span (gain) and $b$ the zero (offset).
Both drift, so a field calibration re-fits the two constants against
two reference gases. Suppose the sensor reports $y_{1} = 455 \,\text
{ppm}$ against an outdoor-ambient reference of $x_{1} = 420 \,\text
{ppm}$ and $y_{2} = 1980 \,\text{ppm}$ against a bottled span gas of
$x_{2} = 2000 \,\text{ppm}$. The two points fix the correction. The
span is $m = (y_{2} - y_{1})/(x_{2} - x_{1}) = (1980 - 455)/
(2000 - 420) = 1525/1580 = 0.965$, and the zero is
$b = y_{1} - m x_{1} = 455 - 0.965 \times 420 = 455 - 405.6 =
49.4 \,\text{ppm}$. The inverse map that the firmware then applies to
a raw reading $y$ is $\hat{x} = (y - b)/m = (y - 49.4)/0.965$. A
subsequent raw reading of $y = 1000 \,\text{ppm}$ corrects to
$\hat{x} = (1000 - 49.4)/0.965 = 985 \,\text{ppm}$, a $1.5\%$
correction that would otherwise be a $50 \,\text{ppm}$ project-budget
violation on the $\pm 50 \,\text{ppm}$ target. The sensitivity of the
corrected value to the span-gas certificate is direct: a span gas
certified to $\pm 2\%$ (that is, $2000 \pm 40 \,\text{ppm}$)
propagates through $m$ as roughly a $\pm 2\%$ uncertainty on the
corrected reading at the high end, or $\pm 20 \,\text{ppm}$ at
$1000 \,\text{ppm}$, which is why a working carbon-dioxide
calibration buys a certified span gas rather than trusting the
factory constants. The automatic-baseline-correction algorithm in
consumer units performs only the single-point zero half of this
calibration, assuming the sensor periodically sees clean outdoor
air near $420 \,\text{ppm}$; it never re-fits the span, so it cannot
catch a gain drift. That is the failure the two-point procedure
exists to prevent \cite{std:ch05-jcgm-100}.

\engterm{Biosensors}. Enzyme- or antibody-functionalised
transducers detect specific biomarkers. The lateral-flow
immunoassay (the working principle of the home pregnancy test
and the SARS-CoV-2 rapid antigen test at \$$3$) is the simplest
biosensor. Surface-acoustic-wave (SAW) biosensors and
quartz-crystal microbalance (QCM) biosensors detect mass changes
of bound biomolecules on the sensor surface. Microelectrode-array
neural recording instruments (Neuralink Telepathy and similar) and
DNA-sequencing instruments (Illumina NovaSeq X Plus at \$$1.25$
million) extend the biosensor category to the systems-level
chemistry of the chapter's last decades.

Chemical sensors share a characteristic challenge: \engterm
{selectivity}. Few chemical sensing principles respond to a
single analyte; most respond to a class of similar species, and
the engineer must either accept the cross-sensitivity (and
report a single concentration as a class equivalent) or supply
chemistry that compensates for it (a Henderson-Hasselbalch buffer
holds the pH against the cross-interference of a $\mathrm{Na}^{+}$
ISE; an ion chromatograph separates the species before
quantification; a mass spectrometer separates by mass-to-charge
ratio for unambiguous identification).

\subsection{Magnetic}

Magnetic sensors translate magnetic field into an electrical
signal. The dominant types span fifteen orders of magnitude in
sensitivity, from teslas to femtoteslas, and three orders of
magnitude in cost.

\engterm{Hall-effect sensors}. A current $I$ flowing through a
thin semiconductor slab in a transverse magnetic field $B$
develops a Hall voltage $V_{H} = R_{H} I B/t$ where $R_{H} = 1/
(n e)$ is the Hall coefficient and $t$ is the slab thickness. The
working sensitivity is set by the carrier concentration $n$:
silicon ($n \sim 10^{16}$ to $10^{19}\,\si{\per\centi\meter\cubed}$
in doped slabs) is the consumer-electronics standard; GaAs and
InSb ($n \sim 10^{15}\,\si{\per\centi\meter\cubed}$) offer
$100\times$ higher sensitivity at higher cost. Specific
instruments: Allegro A1308 ($\$1$ unit price at thousands,
$\pm 100\,\si{\milli\tesla}$ range, $\pm 0.5\%$ accuracy, $7\,
\si{\micro\tesla}$ noise floor over a $10\,\si{\hertz}$ band) for
DC current sensing; Honeywell SS49E linear Hall ($\$2$,
$\pm 100\,\si{\milli\tesla}$ range, $50\,\si{\micro\tesla}$
noise) for position sensing; Asahi Kasei HG-302C precision Hall
($\$50$, $\pm 5\,\si{\micro\tesla}$ noise, integrated thermal
compensation) for geomagnetic and NDT applications. Hall sensors
have moderate temperature sensitivity ($\sim 0.1\%/\si{\kelvin}$
for typical silicon parts) and respond from DC to $\sim 30\,\si{
\kilo\hertz}$.

\paragraph{Worked example: Hall element in a flux-concentrating
current sensor.} A DC-capable current sensor passes the measured
conductor through a slotted soft-iron core and places a Hall
element in the slot, where the core concentrates the conductor's
field. For a conductor carrying $I_{\text{cond}} = 40 \,\si{\ampere}$
through a core that concentrates the field in the slot to
$B = 20 \,\si{\milli\tesla}$ (a typical value for a modest core and
air gap), a Hall element with sensitivity $S_{H} = 30 \,\si{\milli
\volt}$ per $100 \,\si{\milli\tesla}$ (that is, $0.30 \,\si{\volt
\per\tesla}$ at its rated bias current) produces
$V_{H} = S_{H} B = 0.30 \times 0.020 = 6.0 \,\si{\milli\volt}$.
This is the raw output before the on-die amplifier; a sensor such
as the Allegro-class part scales it to a convenient
$V_{\text{out}} = 40 \,\si{\milli\volt\per\ampere}$ at the pin, so
$40 \,\si{\ampere}$ reads $1.6 \,\si{\volt}$ above the quiescent
mid-supply offset. Two error terms dominate the working budget.
First, the Hall element's offset (the residual $V_{H}$ at zero
field, from the misalignment of the voltage contacts) is of order
$50 \,\si{\micro\tesla}$ referred to field, which maps to
$50 \times 10^{-6}/0.020 \times 40 \,\si{\ampere} = 0.1 \,\si{
\ampere}$ of zero error, or $0.25\%$ of the reading. Second, the
temperature coefficient of the Hall sensitivity, $\sim 0.1\%/\degC$,
turns a $30 \,\degC$ rise above the calibration temperature into a
$3\%$ span error, larger than the offset term. The engineering
lesson mirrors the shunt-versus-CT trade-off already seen: the Hall
sensor buys DC capability and galvanic isolation, and pays for it
with a temperature-dependent span that a working design either
compensates with an on-die reference or trims out with a
periodic zero against a known-zero-current condition.

\engterm{Magnetoresistive sensors}. The resistance of a
ferromagnetic structure varies with the applied field.
Anisotropic magnetoresistance (AMR, $\Delta R/R \sim 2\%$ at room
temperature) is the basic effect in NiFe (permalloy) thin films;
giant magnetoresistance (GMR, $\Delta R/R \sim 5$-$20\%$, Nobel
2007 to Albert Fert and Peter Grünberg) exploits spin-dependent
electron scattering in alternating ferromagnetic/non-magnetic
multilayers; tunnel magnetoresistance (TMR, $\Delta R/R \sim 100$-
$600\%$ at room temperature) uses spin-dependent tunnelling
through a thin MgO barrier. Working instruments: NVE AAH002-02E
(AMR, $\pm 1\,\si{\milli\tesla}$ range, $1\,\si{\nano\tesla}$
noise over $1\,\si{\hertz}$, $\$3$) for compass and current-
sensing; Sensitec GMR-C12 ($\pm 4\,\si{\milli\tesla}$, $0.5\,
\si{\nano\tesla}$ noise, $\$15$) for precision position sensing;
NVE TMR2305 ($\pm 50\,\si{\micro\tesla}$, $10\,\si{\pico\tesla}/
\sqrt{\si{\hertz}}$ noise floor in $f > 100\,\si{\hertz}$, \$60)
for digital read heads in hard drives and high-resolution
position encoders. MR sensors approach Hall in cost and exceed
them by $100$ to $10000\times$ in sensitivity.

\engterm{Fluxgate magnetometers}. A saturable ferromagnetic core
is driven into and out of saturation by an excitation winding at
$1$-$10\,\si{\kilo\hertz}$; an external field unbalances the
hysteresis loop, producing a second-harmonic component in the
sense winding's output proportional to the field. Working
sensitivity: $10\,\si{\pico\tesla}/\sqrt{\si{\hertz}}$ for
laboratory-grade units (Bartington Mag-03 at $\$1500$, three-axis
with $0.07$-$0.10\,\si{\degree}$ orthogonality), $1\,\si{\pico
\tesla}/\sqrt{\si{\hertz}}$ for high-end space-qualified units
(Magson FGM-3D for ESA Cluster and Swarm missions). Fluxgate is
the working choice for absolute DC field measurements at the
nanotesla level: geomagnetic surveys, spacecraft attitude
determination, magnetic-anomaly detection (MAD) in
antisubmarine warfare.

\engterm{Optically-pumped atomic magnetometers}. Alkali vapour
(rubidium, caesium, potassium) in a glass cell is optically
pumped to a non-thermal spin state; the precession of the
magnetisation in the ambient field is detected through the
absorption or polarisation rotation of a probe beam. Sensitivity
reaches $1\,\si{\femto\tesla}/\sqrt{\si{\hertz}}$ in spin-
exchange-relaxation-free (SERF) mode at near-zero field
(QuSpin QZFM-Gen3 at $\$60{,}000$, used for magnetoencephalography
arrays). At Earth field, optically pumped scalar magnetometers
(Geometrics G-823A caesium-vapour at $\$15{,}000$, $10\,\si{\pico
\tesla}/\sqrt{\si{\hertz}}$, $0.01\,\si{\nano\tesla}$ absolute
accuracy) replace fluxgate for high-precision geophysical and
archaeological survey.

\engterm{Superconducting quantum interference devices (SQUIDs)}.
A superconducting loop interrupted by one (rf SQUID) or two (dc
SQUID) Josephson junctions converts flux through the loop to a
voltage modulated periodically at one flux quantum $\Phi_{0} =
h/(2e) \approx 2.068 \times 10^{-15}\,\si{\weber}$. Low-T$_{c}$
(niobium, $4.2\,\si{\kelvin}$ operating in liquid helium) SQUIDs
reach $1$-$3\,\si{\femto\tesla}/\sqrt{\si{\hertz}}$ noise floor
(StarCryo dc SQUID at $\$30{,}000$ plus cryogenic infrastructure).
High-T$_{c}$ (YBa$_{2}$Cu$_{3}$O$_{7-x}$, $77\,\si{\kelvin}$
operating in liquid nitrogen) SQUIDs reach $\sim 30\,\si{\femto
\tesla}/\sqrt{\si{\hertz}}$ at $10\times$ lower cryogenic cost.
Working applications: magnetoencephalography (MEG, $\sim 300$-
sensor helmet arrays at hospital MEG facilities like Elekta
Neuromag TRIUX), magnetocardiography, ultra-low-field NMR,
geomagnetic and biomagnetic prospecting, and the standard kilogram
realisation through the Kibble balance (chapter 3 of this
volume).

The engineering choice across the seven magnetic sensor families
spans fifteen orders of magnitude in sensitivity (Hall at $\sim
10\,\si{\micro\tesla}$ noise, SQUID at $\sim 1\,\si{\femto
\tesla}/\sqrt{\si{\hertz}}$). The selection is set by the
required sensitivity, the working bandwidth, the cost budget,
and whether DC capability and absolute calibration are required.
A Hall sensor measures the field of a permanent magnet at
$\$1$; a SQUID measures the magnetic field of a neuron's action
potential through the skull at $\$30{,}000$ plus cryogenic
infrastructure. The intermediate niches (MR, fluxgate, atomic)
each occupy specific cost-sensitivity-bandwidth regions of the
design space.

\section{Sensor specifications: accuracy, precision, resolution,
range, drift, hysteresis}\pathtag{core}

A sensor's datasheet is its public contract. The reader who buys a
sensor on its datasheet should understand what each specification
asserts and what it does not. \Cref{fig:vol01:ch05:spec-anatomy}
shows the terms on a single transfer plot: offset, nonlinearity,
hysteresis, resolution, range, saturation, and drift, each visible
as a separate feature of the actual curve, with bandwidth and
response time deferred to a time-domain view. The GUM
\cite{std:ch05-jcgm-100} is the working framework for combining
each term into a single combined standard uncertainty for the
measurement chain; the chapter project's uncertainty budget in
\repopath{data/sensor_uncertainty_budget.csv} is one worked example.

\input{volumes/01-quantity/ch05-sensors-and-instruments/figures/fig-spec-anatomy.tex}

\subsection{Accuracy and precision}

\engterm{Accuracy} is the closeness of the sensor's output to the
true value of the input quantity. \engterm{Precision} is the
closeness of repeated outputs to one another for the same input. The
distinction was made formally in chapter 1 of this volume; we
restate it here in the sensor's terminology because datasheets
sometimes use the words loosely.

A sensor specification of "$\pm 0.5\,\%$ of reading" without further
qualification is a manufacturer's error limit under stated
conditions, not automatically a coverage interval. The datasheet may
define it as maximum permissible error, expanded uncertainty,
typical error, or a test limit. The reader should not infer
$k = 1$, $k = 2$, or a rectangular distribution from the plus-minus
sign alone. Chapter 4 of this volume warned the reader against
exactly this inference; the warning applies in full to sensor
datasheets.

\subsection{Resolution}

\engterm{Resolution} is the smallest change in the input quantity
that produces a discernible change in the output. For a digital
instrument, resolution is typically the value of the least
significant bit. For an analog instrument, resolution is set by the
display, the noise floor, and the operator's discrimination.

Resolution is not the same as accuracy. A sensor with resolution
$0.01 \,\si{\degreeCelsius}$ may have an absolute accuracy of
$\pm 1 \,\si{\degreeCelsius}$. The reader who reports the
high-resolution figure as if it were the accuracy has overstated
their measurement.

\paragraph{Worked example: kitchen scale.} A digital kitchen scale
displays mass to $1 \,\si{\gram}$ resolution across a $5 \,\si
{\kilogram}$ range. Its accuracy specification, often buried in the
back of the manual, may be of order $\pm 5 \,\si{\gram}$ at full
scale plus $\pm 1 \,\si{\gram}$ at low loads. A reader weighing
$200 \,\si{\gram}$ of flour reads $201 \,\si{\gram}$ on the display
and concludes the flour weighs $201 \,\si{\gram}$ to the nearest
gram; the working accuracy of the measurement is closer to
$\pm 3 \,\si{\gram}$ once linearity, repeatability, and the
display's resolution are combined as a sum of variances. The
display gives one significant digit beyond what the working
accuracy supports. Reporting the mass as $201.0 \,\si{\gram}$
overstates the precision; reporting it as $200 \pm 3 \,\si{\gram}$
states what the instrument actually contains.

\subsection{Range}

\engterm{Range} is the span of input values over which the sensor
produces an output meeting its other specifications. Above the
range, the sensor saturates or is damaged; below the range, the
sensor's output may be dominated by noise or offset.

Range is not always equivalent to "operating range." A pressure
transducer rated for $0$ to $10\,\text{bar}$ may survive a higher
proof or burst pressure, but readings above $10\,\text{bar}$ are
outside the calibrated range. Survival is a mechanical claim;
calibration is a measurement claim.

\subsection{Drift}

\engterm{Drift} is the slow change in a sensor's output for a
constant input. Mechanisms include component aging, surface
contamination, mechanical wear, and environmental cycling. We took
up drift formally in chapter 3 of this volume in the context of
calibration intervals; sensor datasheets typically state drift in
units of "output units per year" or "output units per
$10{,}000 \,\si{\hour}$ of operation."

A sensor's drift is one of the parameters that sets its
recalibration interval. A drift specification of $0.1\,\%$ per year
on a sensor operating in an application that requires $1\,\%$
accuracy gives only a first bound on recalibration interval. The
working interval also depends on initial calibration uncertainty,
drift distribution, guard band, environmental exposure, and
consequence of error. The general engineering rule applies: the
manufacturer's drift specification is one component of the working
uncertainty budget, not a complete characterisation.

\begin{mastery}
A single drift number such as $0.1\%$ per year collapses a
sensor's whole stability behaviour into one figure and hides the
timescale at which averaging stops helping. The
\engterm{Allan deviation} $\sigma_{A}(\tau)$, borrowed from the
frequency-standards community \cite{paper:ch05-allan-1966}, keeps
the timescale. It is the root-mean-square of successive differences
of the sensor's output averaged over blocks of duration $\tau$:
\[
\sigma_{A}^{2}(\tau) = \frac{1}{2} \left\langle
(\bar{y}_{k+1} - \bar{y}_{k})^{2} \right\rangle,
\]
where $\bar{y}_{k}$ is the mean of the output over the $k$-th block
of length $\tau$. Plotted against $\tau$ on log--log axes, the curve
reads as a sequence of noise regimes. On the short-$\tau$ side it
falls with slope $-1/2$: this is white noise, and averaging longer
helps, exactly as $\sigma/\sqrt{N}$ promises. It reaches a minimum,
the \engterm{bias instability}, at some optimum averaging time
$\tau_{\text{opt}}$: this is the best stability the sensor can
deliver, and it is the number a working budget should carry rather
than the raw per-sample noise. Beyond $\tau_{\text{opt}}$ the curve
rises again with slope $+1/2$ (random walk) and then $+1$ (linear
drift): here averaging \emph{hurts}, because the sensor is wandering
faster than the mean can track it. For the smartphone gyroscope of
the exercises, $\tau_{\text{opt}}$ is a fraction of a second and the
minimum is a few tenths of a degree per hour; for a navigation-grade
ring-laser gyro, $\tau_{\text{opt}}$ is minutes and the minimum is
$0.001 \,\si{\degree}/\si{\hour}$. The Allan curve is the honest
answer to the averaging-time question that a single drift number
cannot give: average to $\tau_{\text{opt}}$ and no longer, then
recalibrate against an independent reference. The same curve
governs the recalibration-interval estimate in section~5.4, where
the drift term that consumes the budget is the random-walk rise to
the right of $\tau_{\text{opt}}$.
\end{mastery}

\subsection{Hysteresis}

\engterm{Hysteresis} is the dependence of a sensor's output not
only on the current input but on the history of the input. A
mechanical pressure transducer's reading at $5 \,\si{\bar}$
approached from below may differ from its reading at $5 \,\si{\bar}$
approached from above. The difference, expressed as a percentage of
the full-scale range, is the hysteresis specification.

Hysteresis is unavoidable in many physical sensing principles
(magnetic, mechanical, elastic). The discipline that limits its
effect is to ensure measurements are taken at consistent approach
direction, or to record the approach direction as part of the
measurement.

\subsection{Other specifications}

A reader buying a sensor will encounter additional specifications,
each addressing a separate failure mode of an idealised sensor:

\begin{itemize}[leftmargin=*]
  \item \engterm{Linearity}: deviation of the output from a
    straight-line fit to its calibration curve.
  \item \engterm{Repeatability}: closeness of repeated outputs to
    the same input under identical conditions, typically across
    short timescales.
  \item \engterm{Reproducibility}: closeness of outputs to the same
    input across different laboratories, days, or operators.
  \item \engterm{Bandwidth}: the frequency range over which the
    sensor responds to time-varying inputs without significant
    attenuation or phase shift.
  \item \engterm{Response time}: the time required for the sensor's
    output to reach a specified fraction (typically $63\,\%$ or
    $90\,\%$) of its final value following a step change in input.
  \item \engterm{Cross-sensitivity}: the sensor's response to
    quantities other than the intended measurand (a temperature
    sensor that responds to humidity, a strain gauge that responds
    to temperature).
  \item \engterm{Operating environment}: temperature range, humidity
    range, vibration tolerance, shock tolerance, electromagnetic
    interference (EMI) tolerance.
\end{itemize}

A reader who selects a sensor without reading its operating
environment specification has chosen a sensor that may not work in
its intended location. We come back to selection in section 5.4.

\section{Choosing the right sensor}\pathtag{core}

Sensor selection starts as a checklist and becomes judgment with
use. The checklist matters because it forces the measurand,
environment, interface, cost, and calibration record into the same
decision.

\subsection{The procedure}

\begin{enumerate}[leftmargin=*]
  \item \engterm{State the measurand}. The first step is the
    one most often skipped. The reader should write down what is
    being measured, in what units, with what required uncertainty.
    A vague specification ("we need to measure the temperature") is
    inadequate; a precise specification ("the surface temperature
    of the test article at five locations, sampled at $1 \,\si
    {\hertz}$, with combined standard uncertainty better than
    $0.5 \,\si{\degreeCelsius}$, in the range $0$ to $80 \,\si
    {\degreeCelsius}$") is what selection requires.
  \item \engterm{Identify candidate sensing principles}. Given the
    measurand and required performance, the section 5.2 catalogue
    narrows the field. For temperature in the stated range, RTDs
    and thermistors are the standard candidates; thermocouples are
    available but require reference junction handling.
  \item \engterm{Survey commercial offerings}. For each candidate
    principle, identify two to four commercially available sensors
    that meet the specifications. Read the datasheets in full,
    including the small print on operating range, environmental
    limits, and conditions of accuracy.
  \item \engterm{Compare specifications against requirements}.
    Construct a table comparing each candidate against the
    measurand specification. Identify which specifications are
    most marginal for each candidate.
  \item \engterm{Account for environmental constraints}. The
    physical environment of the application (temperature, humidity,
    vibration, EMI, chemical exposure) eliminates many candidates.
    A perfect-on-the-bench sensor that fails in field conditions is
    a failed selection.
  \item \engterm{Account for integration constraints}. The sensor
    must connect to the measurement system that uses its output.
    Signal conditioning, ADC interface, mechanical mounting, and
    cable management all enter the selection.
  \item \engterm{Account for cost and supply}. Working sensors are
    sometimes selected by what is in stock; working engineers
    sometimes accept a slightly inferior sensor to avoid a
    six-month lead time. The selection record should name these
    trade-offs explicitly.
  \item \engterm{Document the selection}. The selection record is
    part of the project's design documentation; it survives the
    engineer's tenure and informs later substitutions when the
    selected sensor is discontinued.
\end{enumerate}

\Cref{fig:vol01:ch05:selection-tree} traces the eight steps as a
flowchart with the working return path made explicit: a candidate
eliminated at step 5 (environment) or step 6 (integration) sends the
engineer back to step 2 (principles), not to step 3
(survey), because the principle catalogue is where alternative
candidates live. The IEEE TEDS standard
\cite{std:ch05-ieee-1451-2-1997} formalises one downstream benefit of
documenting the selection: a sensor whose calibration coefficients
and unit information are carried inside the sensor itself (a
"plug-and-play" or smart-transducer interface) reduces the chance
that a later substitution carries the wrong coefficients into the
host instrument.

\input{volumes/01-quantity/ch05-sensors-and-instruments/figures/fig-selection-tree.tex}

\subsection{The thermistor case}

The reader's project in this chapter selects an NTC thermistor for
a household-temperature measurement application. The selection is
straightforward because the application is undemanding: we want to
measure water and air temperatures in a kitchen environment over
the range $0$ to $80 \,\si{\degreeCelsius}$ with expanded
uncertainty no larger than $\pm 1 \,\si{\degreeCelsius}$ at a
stated coverage factor. The lower range keeps the project inside
the settled low-hazard brief.

For this application, a $10 \,\si{\kilo\ohm}$ NTC thermistor (B
parameter $\sim 3950 \,\si{\kelvin}$) read through a voltage divider
into a microcontroller's $10$-bit or $12$-bit ADC is a defensible
choice. The thermistor's strong temperature dependence (resistance
ratio of about $40$ across the working range) makes it sensitive
enough that ADC quantisation is rarely the dominant uncertainty for
a one-degree project. The nonlinearity is captured by the
Steinhart-Hart equation:

\[
\frac{1}{T} = A + B \ln\!\left(\frac{R}{R_{\ast}}\right) + C
\left[\ln\!\left(\frac{R}{R_{\ast}}\right)\right]^{3},
\]

where $T$ is absolute temperature in kelvin, $R$ is the thermistor
resistance, $R_{\ast}$ is a reference resistance (commonly
$1 \,\si{\ohm}$ when handbook coefficients are reported for the
numerical value of $R$ in ohms), and $A, B, C$ are coefficients
fitted from a small set of calibration measurements. The logarithm
must act on a dimensionless ratio or on a stated numerical
convention; the chapter takes up dimensional discipline in
chapter 2 of this volume \cite{text:fraden2016}. For a less
stringent application, the $\beta$-parameter approximation,

\[
R = R_{0} \exp\!\left(\beta \left(\frac{1}{T} - \frac{1}{T_{0}}
\right)\right),
\]

is often adequate; it is the form the project uses.

The chosen sensing principle would not work for the same range with
a stringent uncertainty requirement, in a high-vibration
environment, or with a one-second response-time requirement. A
different application would lead to a different selection.
\Cref{fig:vol01:ch05:thermistor-divider} shows the schematic and
the transfer curve $V_{\text{out}}(T)$ together; the slope is
largest at the symmetric operating point where
$R_{\text{th}} \approx R_{\text{ref}}$, which sets the working
ADC-quantisation budget. The script
\repopath{code/thermistor_divider.py} reproduces the curve and
locates the temperature of maximum sensitivity numerically;
\repopath{code/adc_noise_simulation.py} adds one-LSB Gaussian noise
on a 10-bit ADC and reports the standard deviation of the inferred
temperature at four points across the range.

\input{volumes/01-quantity/ch05-sensors-and-instruments/figures/fig-thermistor-divider.tex}

\begin{estimation}
\textbf{Estimate the analog-to-digital converter resolution required
for the thermistor measurement.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
Define the divider before estimating it. The fixed reference
resistor is the upper leg from $V_{\text{cc}}$ to the ADC node, and
the thermistor is the lower leg from the ADC node to ground, so the
midpoint voltage is

\[
V_{\text{out}} = V_{\text{cc}} \frac{R_{\text{th}}}{R_{\text{ref}}
+ R_{\text{th}}}.
\]

With a $10 \,\si{\kilo\ohm}$ reference resistor and a $10 \,\si
{\kilo\ohm}$ thermistor at room temperature, the midpoint sits near
half the supply.

For a one-degree project with $V_{\text{cc}} = 3.3 \,\si{\volt}$ and
a $10$-bit ADC, one LSB corresponds to roughly
$0.1 \,\si{\degreeCelsius}$ near $25 \,\si{\degreeCelsius}$ and
several times that near the ends of the working range, so a $10$-bit
ADC is adequate near room temperature and marginal at the extremes.
A $12$-bit ADC gives sub-degree resolution across the range. ADC
quantisation is usually smaller than the reference thermometer
uncertainty, the resistor tolerance, electrical noise, self-heating,
and fit residuals for a one-degree project. A common through-hole
resistor may be $\pm 5\,\%$; ordinary metal-film parts are often
$\pm 1\,\%$; precision reference resistors may be $\pm 0.1\,\%$ or
better. A reader specifying this measurement chain knows where to
invest precision budget: a precision reference resistor and a
calibrated reference thermometer for the calibration session, not a
$16$-bit ADC.
\end{estimation}

\begin{estimation}
\textbf{Estimate the recalibration interval for a household-grade
sensor on a one-degree budget.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A representative consumer-grade NTC thermistor and its conditioning
chain shows long-term drift of order $0.1 \,\%$ of reading per year
under sealed indoor service. A $10 \,\si{\kilo\ohm}$ thermistor read
at $25 \,\degC$ drifts by approximately $10 \,\si{\ohm}$ per year;
through the beta-parameter model with
$\beta = 3950 \,\si{\kelvin}$, the apparent temperature change is
approximately
\[
\Delta T \;\approx\; \frac{T^{2}}{\beta} \frac{\Delta R}{R}
\;\approx\; \frac{(298)^{2}}{3950} \cdot 10^{-3}
\;\approx\; 0.022 \,\degC \text{ per year}.
\]
On a $\pm 1 \,\degC$ project budget, drift alone consumes about
$2\,\%$ of the budget per year, and a recalibration every five
years (about $0.1 \,\degC$ accumulated) is comfortable; on a
$\pm 0.1 \,\degC$ project budget, drift consumes about $20\,\%$ of
the budget per year, and an annual recalibration is the working
choice. The estimate uses one term (drift); a complete recalibration
interval also accounts for guard band, environmental cycling, and
consequence of error \cite{std:ch05-jcgm-100}. The Allan-variance
literature \cite{paper:ch05-allan-1966} provides the working
stability statistic for long-term drift studies in precision
applications. Compare with chapter 3 of this volume, where the
recalibration-interval principle was first introduced; the present
estimate is the per-application instantiation.
\end{estimation}

\subsection{Three worked selections}

The thermistor case is the project's own selection and is
deliberately undemanding. Three further selections, each
representative of a class of measurement the reader will encounter,
work the procedure end to end against named commercial parts. Each
follows the eight-step flowchart of
\cref{fig:vol01:ch05:selection-tree}, names the candidates
considered, and documents the lifecycle cost.

\paragraph{Selection 1: water temperature in a heating loop.}
A residential or light-industrial hydronic heating loop circulates
water from a boiler through radiators or radiant-floor manifolds and
back. We want to measure the loop temperature near the boiler outlet
to feed a closed-loop controller. The design constraints are stated
before any sensor is named:

\begin{itemize}[leftmargin=*]
  \item \emph{Measurand}: water temperature, single point near the
    boiler outlet, sampled at $1\,\si{\hertz}$.
  \item \emph{Range}: $0$ to $95 \,\degC$ (covering cold start through
    boiler high limit).
  \item \emph{Accuracy}: combined standard uncertainty
    $\pm 0.5 \,\degC$ across the range.
  \item \emph{Response time}: $\tau_{63} < 2 \,\si{\second}$ to detect
    a stall or runaway within one controller period.
  \item \emph{Form factor}: immersion in a stainless 1/4-inch
    thermowell, $50 \,\si{\milli\meter}$ deep, with $1/4\,\text{in}$
    NPT fitting.
  \item \emph{Cost}: $\$50$ to $\$200$ per unit, including probe
    and conditioning.
  \item \emph{Service life}: $10$ years sealed; no field calibration.
\end{itemize}

Step 2 (principle catalogue). For temperature in the $0$ to
$100 \,\degC$ band, the working candidates are NTC thermistor,
platinum RTD, and type-T or type-K thermocouple. We rule each in or
out by the constraints. A bead thermistor in a stainless sheath
reaches $\tau_{63} \approx 1$ to $3 \,\si{\second}$, accuracy
$\pm 0.2 \,\degC$ for a $\beta$-fit consumer part; defensible but
each manufacturer's coefficients drift over $10$ years of $90 \,\degC$
service. A type-T thermocouple in a $1/4 \,\text{in}$ sheath
reaches $\tau_{63} \approx 0.5 \,\si{\second}$, accuracy
$\pm 0.5 \,\degC$ Class 1 per IEC 60584-1:2013, and requires cold-
junction compensation in the conditioning electronics. A Class A
Pt100 RTD per IEC 60751:2022 reaches $\tau_{63} \approx 5 \,\si{
\second}$ for a $3 \,\si{\milli\meter}$-diameter sheath in still
water, $\tau_{63} \approx 1.5 \,\si{\second}$ in flowing water, and
$\pm 0.15 \,\degC$ at $0 \,\degC$ with $\pm 0.0035 \cdot \lvert T
\rvert$ slope, well inside the $\pm 0.5 \,\degC$ budget.

Step 3 (commercial offerings). For each candidate principle we
identify two to four parts. For Pt100 Class A we shortlist the
Omega PR-22-3-100-1/4-2-E-T ($\$110$ as of 2026,
$3$-wire, $1/4 \,\text{in}$ sheath, $50 \,\si{\milli\meter}$ length,
hermetic compression seal) \cite{text:ch05t3-omega-pr22-datasheet};
the Heraeus PT100-D-3-100-MS-TC ($\$95$, Class A, $3$-wire,
similar form factor); and the Watlow F-Series Pt100 in a $3 \,\si
{\milli\meter}$ sheath ($\$140$). For thermistor we shortlist a
US Sensor PR222J2 in stainless probe ($\$45$, $\pm 0.2 \,\degC$
interchangeability, $\beta = 3892$); for thermocouple, an Omega
TJ36-CASS-18U-12 type-T ($\$60$, $1.6 \,\si{\milli\meter}$
sheath, ungrounded).

\begin{table}[ht]
\centering
\caption[Decision matrix: water-temperature selection]{Decision
matrix for the heating-loop selection. Specifications drawn from
manufacturer datasheets current as of 2026. The Pt100 Class A wins
on accuracy headroom and ten-year service life; the type-T wins on
response time alone; the thermistor wins on price but loses on
long-term drift through repeated $90 \,\degC$ excursions.}
\label{tab:vol01:ch05:selection-water}
\begin{tabular}{lccc}
\toprule
Criterion & Pt100 Class A & Type-T TC & NTC Thermistor \\
\midrule
Accuracy ($0 \,\degC$)            & $\pm 0.15 \,\degC$ & $\pm 0.5 \,\degC$ & $\pm 0.2 \,\degC$ \\
Accuracy ($90 \,\degC$)           & $\pm 0.47 \,\degC$ & $\pm 0.5 \,\degC$ & $\pm 0.5 \,\degC$ \\
Response time $\tau_{63}$ (water) & $\sim 1.5 \,\si{\second}$ & $\sim 0.5 \,\si{\second}$ & $\sim 1$ to $3 \,\si{\second}$ \\
10-year drift                     & $< 0.05 \,\degC$ & $\pm 0.2 \,\degC$ & $\pm 0.3 \,\degC$ \\
Probe cost                        & $\$110$ & $\$60$ & $\$45$ \\
Conditioning cost                 & $\$40$ (3-wire) & $\$25$ (CJC IC) & $\$10$ (divider) \\
Total (probe + cond.)             & $\$150$ & $\$85$ & $\$55$ \\
Lifecycle 10 y (recal.)           & $\$200$ (one recal.) & $\$285$ (yearly recal.) & $\$155$ (replace) \\
\bottomrule
\end{tabular}
\end{table}

Step 8 (selection). The selected sensor is the Omega
PR-22-3-100-1/4-2-E-T Class A Pt100. The decision record states:
"Pt100 Class A IEC 60751:2022 chosen for accuracy headroom across
the working range, ten-year drift well inside the $\pm 0.5 \,\degC$
budget, and integration with the building-management system's
standard $3$-wire Pt100 input card. Type-T thermocouple was the
closest alternative on response time but loses on $10$-year drift
through cold-junction-compensation IC ageing. Thermistor was the
lowest-cost alternative but loses on accumulated drift through
repeated $90 \,\degC$ thermal cycling." Lifecycle cost over ten
years (probe $\$110$, signal-conditioning card share $\$40$, one
recalibration at year five against a class A reference for $\$50$
plus shipping) totals approximately $\$200$. The thermistor's
apparent $\$55$ saving is consumed by replacement at year five plus
the cost of a second uncertainty budget covering the substitution.

\paragraph{Selection 2: vibration on a rotating shaft.}
A $1500 \,\text{rpm}$ centrifugal pump shaft is being monitored for
incipient bearing failure. We want to instrument an axial location
on the bearing housing with a single-axis accelerometer.

\begin{itemize}[leftmargin=*]
  \item \emph{Measurand}: housing vibration, single-axis (radial), in
    units of $g$ at the bearing housing.
  \item \emph{Frequency band}: $10 \,\si{\hertz}$ to $1 \,\si{\kilo
    \hertz}$ (shaft rotation $25 \,\si{\hertz}$, blade-pass
    harmonics through $400 \,\si{\hertz}$, bearing-defect
    frequencies through $1 \,\si{\kilo\hertz}$).
  \item \emph{Amplitude range}: $\pm 50 \,g$ peak (bearing-defect
    impulses).
  \item \emph{Environment}: housing surface temperature $60 \,\degC$,
    dust, occasional water washdown.
  \item \emph{Mounting}: removable magnet base for periodic spot
    measurement; permanent stud-mount option held in reserve.
  \item \emph{Signal interface}: $4 \,\si{\milli\ampere}$
    constant-current ICP (IEPE) for compatibility with the existing
    data-acquisition system.
  \item \emph{Cost}: $\$300$ to $\$800$ per channel, sensor plus
    conditioner.
\end{itemize}

Step 2 (principle catalogue). Piezoelectric ICP accelerometers
dominate this band and amplitude range. MEMS accelerometers reach
$\pm 50 \,g$ but show higher $1/f$ noise and a typical low-frequency
cutoff at $0.5 \,\si{\hertz}$ that meets the band but at degraded
SNR. Capacitive accelerometers are ruled out by the $1 \,\si{\kilo
\hertz}$ upper band requirement (typical cutoff $400 \,\si{\hertz}$).

Step 3 (commercial offerings). For piezoelectric ICP we shortlist
the PCB Piezotronics 352A24 ($100 \,\si{\milli\volt}/g$, $0.8 \,\si
{\hertz}$ to $10 \,\si{\kilo\hertz}$ $\pm 5\,\%$ band, $\pm 500 \,g$
peak, $4.5 \cdot 10^{-3} \,g/\sqrt{\si{\hertz}}$ noise floor,
$\$425$) \cite{text:ch05t3-pcb-352a24-datasheet}; the IMI 603C01
($100 \,\si{\milli\volt}/g$, $0.5 \,\si{\hertz}$ to $10 \,\si{\kilo
\hertz}$, $\$330$); and the Endevco 7259B
($100 \,\si{\milli\volt}/g$, $0.5 \,\si{\hertz}$ to $5 \,\si{\kilo
\hertz}$, $\$650$).

Step 5 (environment). The $60 \,\degC$ housing temperature is well
inside the $-54$ to $+121 \,\degC$ operating range of all three.
Dust and washdown rule out the IMI 603C01's nominally rated IP65
enclosure in practice; the PCB 352A24's hermetic seal carries IP68
and clears the washdown event without question. The Endevco 7259B
is also hermetic but uses a $10$-$32$ coaxial connector that is
slower to remove and refit during periodic measurement.

Step 6 (integration). The existing data-acquisition system supplies
$4 \,\si{\milli\ampere}$ through a PCB 482C04 four-channel signal
conditioner \cite{text:ch05t3-pcb-482c-datasheet}; all three
candidates accept the standard $4 \,\si{\milli\ampere}$ ICP excitation.
A two-pin (BNC or $10$-$32$ on the sensor side) magnet base
(PCB 080A130, $\$95$) brings the total per-measurement
channel to roughly $\$520$ for the PCB pair (sensor plus magnet
base; the conditioner is amortised across four channels).

Step 8 (selection). The PCB 352A24 with a PCB 080A130 magnet base,
fed to a PCB 482C04 channel of the existing acquisition system. The
decision record states: "PCB 352A24 chosen for hermetic seal under
periodic washdown, mid-band noise floor adequate for the bearing-
defect frequencies of interest, and connector compatibility with
the existing PCB signal-conditioning rack. Endevco 7259B was a
defensible alternative at higher cost but loses on connector
serviceability." Lifecycle cost is dominated by the
signal-conditioning rack, which the plant already owns; the
marginal cost of the new channel is the $\$425$ sensor plus the
$\$95$ base, $\$520$ total.

\paragraph{Selection 3: visible-light intensity in a dim space.}
A photographic-darkroom safelight intensity is being verified across
$10^{-3}$ to $10^{2} \,\si{\lux}$ to confirm safe exposure margins
for orthochromatic film during processing.

\begin{itemize}[leftmargin=*]
  \item \emph{Measurand}: photometric illuminance, in lux, on a
    flat surface oriented toward the safelight.
  \item \emph{Range}: $10^{-3}$ to $10^{2} \,\si{\lux}$ ($5$ decades).
  \item \emph{Accuracy}: $\pm 10\,\%$ of reading across the range,
    traceable to NIST photometric standards.
  \item \emph{Response time}: not load-bearing; readings are taken
    over seconds with the safelight on continuously.
  \item \emph{Spectral response}: photometric ($V(\lambda)$
    correction to the CIE 1931 standard observer).
  \item \emph{Cost}: $\$1{,}000$ to $\$3{,}000$ for the photodiode
    plus traceable calibration certificate.
\end{itemize}

Step 2 (principle catalogue). Working photometric candidates are:
silicon photodiode with $V(\lambda)$-correction filter; CdS
photoresistor; photomultiplier tube (PMT); single-photon avalanche
diode (SPAD). The CdS photoresistor is ruled out for two
reasons: response-time $\tau_{63}$ of order $50 \,\si{\milli\second}$
is acceptable, but the cell's photoconductive response is
nonlinear by $\sim 30\,\%$ across the working decade and shows
strong temperature dependence; and the spectral peak at $530 \,\si
{\nano\meter}$ is uncorrected to $V(\lambda)$, with a peak-response
offset that biases readings of safelights with strong red content.
The PMT is overkill at $10^{2} \,\si{\lux}$: at the top of the
range, the photocathode current would exceed the working linear
range without aggressive neutral-density filtering, defeating the
range advantage.

Step 3 (commercial offerings). The Hamamatsu S2386-18K silicon
photodiode with photometric-correction filter
\cite{text:ch05t3-hamamatsu-s2386-datasheet} reaches dark current
$\sim 50 \,\si{\pico\ampere}$ at $10 \,\si{\milli\volt}$ reverse
bias and NEP $\sim 2 \cdot 10^{-14} \,\si{\watt}/\sqrt{\si{\hertz}}$,
enough for the bottom of the range with a low-noise transimpedance
amplifier. The Hamamatsu S1133-14 ($\$220$) is the
non-photometric counterpart; the Thorlabs S121C ($\$960$
photometric handheld) is a complete instrument with built-in
$V(\lambda)$ correction and a NIST-traceable calibration; the
International Light IL1700 with SED033 detector ($\$2{,}500$,
photometric, NIST-traceable) is the laboratory-grade alternative.

Step 8 (selection). The Thorlabs S121C handheld photometer plus the
Hamamatsu S2386-18K bare diode as a continuous monitor for the
in-darkroom display. The decision record states: "Thorlabs S121C
chosen as the calibrated reference; Hamamatsu S2386-18K chosen as
the in-darkroom continuous monitor and verified against the
S121C at three points across the working range during commissioning.
CdS photoresistor was ruled out for nonlinearity and uncorrected
spectral response; PMT was ruled out for the upper-range
photocathode-current saturation that would force neutral-density
filtering."  Lifecycle cost: $\$960$ for the Thorlabs photometer
(NIST-traceable, six-year recalibration interval at $\$280$ per
cycle), plus $\$160$ for the Hamamatsu diode, plus $\$80$ for
the transimpedance amplifier (a single op-amp transimpedance with
$10^{8} \,\si{\ohm}$ feedback). The total commissioning budget is
roughly $\$1{,}200$, with a single mid-life recalibration of the
S121C at year six adding $\$280$.

The pattern across the three selections is the same. The
measurand specification arrives first. The principle catalogue
narrows the field. The commercial offerings make the field
concrete. The environment and integration eliminate
perfect-on-the-bench candidates that cannot survive the field. The
decision record commits the trade-offs to writing. A reader who
runs the procedure has a working selection at the end; a reader who
skips a step has a candidate, not a selection.

\section{Common instruments: multimeter, oscilloscope, micrometer,
scale, microscope, sensor arrays}\pathtag{standard}

A sensor produces a signal; an instrument turns the signal into an
indication a working engineer can read, record, calibrate, and
challenge. Six instruments form the minimum vocabulary here: the
digital multimeter, oscilloscope, micrometer, scale, microscope,
and sensor array. Volume VIII chapter 11 takes them up at depth in
the design context; the catalogue here is the recognition step.

\subsection{The digital multimeter (DMM)}

A \engterm{multimeter} measures voltage, current, and resistance,
plus capacitance and frequency in modern instruments. A
four-and-a-half-digit handheld DMM and a six-and-a-half-digit
benchtop DMM can differ by orders of magnitude in DC-voltage
accuracy, input burden, stability, and calibration interval.
Numerical comparison should be drawn from cited representative
manufacturer datasheets at a stated date; this chapter does not pin
specific numbers because they age too quickly to live in the
permanent text.

Working principles inside the DMM include voltage-to-frequency
conversion or sigma-delta ADCs for the voltage measurement, current
shunts or transformer coupling for current, and four-wire
(Kelvin) measurement for resistance to eliminate lead resistance.
The reader selecting a DMM for a measurement should consult the
"specifications" page of the datasheet and the "operating ranges"
page; the front-panel display sometimes overstates the available
precision.

\paragraph{Reading the DMM datasheet.} The DC-voltage accuracy
specification on a benchtop DMM is written as a sum of two
fractional terms, with the Keysight 34470A datasheet
\cite{text:ch05t3b-keysight-34470a-datasheet} as the representative
example. The $34470A$'s DC-volts accuracy at the $10 \,\si{\volt}$
range for the $24$-hour, $90$-day, and $1$-year horizons reads as
$\pm (\text{ppm of reading} + \text{ppm of range})$, with the
$24$-hour figure $11 + 0.5$, the $90$-day figure $18 + 0.5$, and the
$1$-year figure $24 + 0.5$, all relative to a calibration temperature
of $23 \pm 5 \,\degC$ after a $90$-minute warm-up. The first term
scales with the input; the second term is a fixed contribution
independent of the input that captures the zero-offset and
gain-error of the instrument. A reading of $1.0 \,\si{\volt}$
on the $10 \,\si{\volt}$ range at the $24$-hour spec carries
$\pm (11 \,\si{ppm} \times 1.0 \,\si{\volt} + 0.5 \,\si{ppm} \times
10 \,\si{\volt}) = \pm (11 + 5) \,\si{\micro\volt} = \pm 16
\,\si{\micro\volt}$; a reading of $10.0 \,\si{\volt}$ on the same
range carries $\pm (110 + 5) \,\si{\micro\volt} = \pm 115 \,\si{
\micro\volt}$. The range term dominates for small readings; the
reading term dominates near full scale. The working engineer's habit
of asking which term dominates the budget is the practical output
of reading the datasheet.

The $24$-hour, $90$-day, and $1$-year horizons capture the
calibration interval. A laboratory that recalibrates its $34470A$
every $90$ days reads the $90$-day spec; one that recalibrates
annually reads the $1$-year spec; one that recalibrates only when
the measurement disagrees with a working reference reads no
published spec at all and must establish its own uncertainty from
check standards in the laboratory. The $24$-hour figure applies
within a calibration session, where the operator has just
transferred a voltage reference and the instrument's zero-offset
has not had time to drift. The three figures form a ladder: the
closer the measurement to the calibration in time, the tighter the
accuracy. An engineer who reports a $24$-hour specification on a
measurement taken six months after calibration has misread the
datasheet.

The temperature coefficient line is the third term. The
$34470A$ specifies $\pm (\text{1 ppm of reading} + 0.1 \,\si{ppm} \,
\text{of range})$ per $\degC$ outside the $23 \pm 5 \,\degC$
calibration window. A laboratory at $30 \,\degC$, $2 \,\degC$
outside the calibration window, adds
$\pm (2 + 0.2) \,\si{ppm}$ to the reading-term and range-term
accuracies. The display carries a digit, the indication-digit,
that is the LSB of the display and not part of the accuracy; the
$34470A$ at the $10 \,\si{\volt}$ range displays to $7.5$ digits
(the indication-digit being the half-digit), with an accuracy of
$\pm 16 \,\si{\micro\volt}$ on a $1.0 \,\si{\volt}$ reading. The
digit beyond the accuracy is a hint about which way the
instrument's zero-offset and gain-error are drifting between
calibrations; it is not a measurement, and reporting it as one
misuses the datasheet.

A worked uncertainty calculation. A laboratory at $25 \,\degC$ reads
$10.0000 \,\si{\volt}$ on a $10 \,\si{\volt}$ DC source. The
laboratory has the $24$-hour spec available because it has just
finished a calibration transfer from a Fluke 732C primary
reference. The instrument-spec contribution:
$\pm (11 + 0.5) \,\si{ppm} \times 10 \,\si{\volt}
= \pm 115 \,\si{\micro\volt}$.
The temperature-coefficient contribution at $\Delta T = 2 \,\degC$:
$\pm (2 \times 1 + 2 \times 0.1) \,\si{ppm} \times 10 \,\si{\volt}
= \pm 22 \,\si{\micro\volt}$.
Combined in quadrature (assuming the two contributions are
independent): $\sqrt{115^{2} + 22^{2}} = 117 \,\si{\micro\volt}$,
giving $\pm 117 \,\si{\micro\volt}$ at the $1\sigma$ level, or
$\pm 234 \,\si{\micro\volt}$ at the $2\sigma$ coverage of
chapter 3. The reportable measurement is
$10.0000 \,\si{\volt} \pm 0.0002 \,\si{\volt}$ at $2\sigma$, with
the source of the uncertainty broken down by component for the
laboratory record. The datasheet has done the chapter-3 calibration
arithmetic for the laboratory; reading it is part of the
measurement.

\subsection{The oscilloscope}

An \engterm{oscilloscope} displays voltage versus time. Modern
digital storage oscilloscopes (DSOs) sample the input at rates
from $\sim 100\,\text{MS/s}$ to $\sim 256\,\text{GS/s}$ with
vertical resolution typically $8$ to $16$ bits. Six specifications
distinguish working instruments.

\engterm{Bandwidth.} The frequency at which the input amplitude
falls to $-3\,\si{\decibel}$ (factor $0.707$). A $100\,\si{\mega
\hertz}$ oscilloscope reads a $100\,\si{\mega\hertz}$ sinusoid at
$70\%$ of its true amplitude; the same scope reads a $10\,\si{
\mega\hertz}$ sinusoid faithfully. For digital signals, choose
oscilloscope bandwidth at least $5\times$ the signal's highest
frequency content (the working rule of thumb captures the fifth
and seventh harmonics of square-wave transitions). The
$5\times$ rule on a $20\,\si{\mega\hertz}$ digital clock requires
$\geq 100\,\si{\mega\hertz}$ scope bandwidth; on a $1\,\si{\giga
\hertz}$ DDR4 memory bus, $\geq 5\,\si{\giga\hertz}$.

\engterm{Sample rate.} For sinusoidal accuracy, sample rate
$\geq 4\times$ bandwidth; for transient capture and accurate
rise-time measurement, $\geq 10\times$ bandwidth. A $1\,\si{
\giga\hertz}$ bandwidth oscilloscope with $5\,\text{GS/s}$
sample rate provides $5$ samples per nominal-bandwidth cycle, the
minimum for sinusoidal interpolation; the same bandwidth with
$10\,\text{GS/s}$ supports transient-edge measurement.

\engterm{Vertical resolution.} An $8$-bit ADC delivers $256$
discrete levels across the input range; $10$-bit $1024$; $12$-bit
$4096$; $16$-bit $65{,}536$ (high-resolution oscilloscopes, HRO).
For a $\pm 5\,\si{\volt}$ input range, an $8$-bit scope quantises
to $39\,\si{\milli\volt}$ per LSB; a $12$-bit scope to $2.4\,\si{
\milli\volt}$; a $16$-bit scope to $153\,\si{\micro\volt}$. HRO
scopes (LeCroy WaveSurfer 3000z with $16$-bit Dual ADC, R\&S
RTM3000 with HD mode at $14$-bit, Pico Technology PicoScope 4000
series at $16$-bit) trade bandwidth for resolution; they are the
working choice for precision-analog measurements where
quantisation noise would otherwise dominate.

\engterm{Triggering.} The scope captures a single waveform when a
trigger condition is met: edge (most common), pulse width
(catching glitches narrower than a specified width), runt (a
pulse that fails to cross a specified threshold), pattern
(combinational logic conditions across multiple channels),
serial-protocol (UART, SPI, I$^{2}$C, CAN, USB, Ethernet,
PCI Express). Modern scopes offer dozens of trigger types;
the user reads the manual or finds an example before the
measurement.

\engterm{Probe.} The probe and its compensation are part of the
measurement chain. The standard $10\times$ passive probe
(Tektronix TPP1000 at \$$220$, Keysight N2862B at \$$190$,
both rated $1\,\si{\giga\hertz}$) presents $10\,\si{\mega\ohm}$
input impedance shunted by $\sim 10\,\si{\pico\farad}$, dividing
the signal by ten and presenting a high-impedance load. At higher
bandwidths, the $\sim 10\,\si{\pico\farad}$ shunt capacitance
dominates ($Z = 1/(2\pi f C) = 16\,\si{\kilo\ohm}$ at $1\,\si{
\giga\hertz}$), loading sensitive circuits. Active probes
(Tektronix TAP1500 at \$$2000$, Keysight N2796A at \$$2500$)
present $0.6$-$1\,\si{\pico\farad}$ shunt capacitance and a
high-input-impedance buffer; the working choice for
multi-gigahertz signal measurement. Differential probes
(Keysight N2750A at \$$5000$) measure between two nodes
referenced to a common-mode rejection; the working choice for
power-electronics and floating-rail measurements.

\engterm{Aliasing and anti-aliasing.} Sampling at $f_{s}$
correctly represents frequencies below $f_{s}/2$ (Nyquist).
Frequencies above $f_{s}/2$ alias back into the displayed
spectrum as artefacts. A $5\,\text{GS/s}$ scope aliases a $3\,
\si{\giga\hertz}$ input to $2\,\si{\giga\hertz}$ on the display.
Working scopes include analog bandwidth limiting (typically a
single-pole low-pass filter at the front end) that attenuates
above the rated bandwidth, but the filter's rolloff at
$f_{s}/2$ is typically only $\sim 20\,\si{\decibel}$, so
aliased components are attenuated but not eliminated.

\input{volumes/01-quantity/ch05-sensors-and-instruments/figures/fig-scope-risetime.tex}

\paragraph{Worked example: choosing scope bandwidth from a rise
time.} A digital designer must measure the edge of a logic signal
whose true $10$--$90\%$ rise time is $t_{r,\text{true}} = 2 \,\si{
\nano\second}$, and asks what oscilloscope bandwidth keeps the
measured rise time within $5\%$ of the truth. A single-pole
front end has a rise-time-bandwidth product of $t_{r,\text{scope}}
\times \text{BW} \approx 0.35$, so a scope of bandwidth
$\text{BW}$ contributes its own rise time
$t_{r,\text{scope}} = 0.35/\text{BW}$. The scope and the signal
add in quadrature: $t_{r,\text{meas}} = \sqrt{t_{r,\text{true}}^{2}
+ t_{r,\text{scope}}^{2}}$. Requiring $t_{r,\text{meas}} \leq
1.05 \, t_{r,\text{true}}$ gives $t_{r,\text{scope}} \leq
t_{r,\text{true}} \sqrt{1.05^{2} - 1} = 2 \,\si{\nano\second}
\times 0.32 = 0.64 \,\si{\nano\second}$, hence
$\text{BW} \geq 0.35/0.64 \,\si{\nano\second} \approx 550 \,\si{
\mega\hertz}$. The signal's own $-3\,$dB bandwidth is
$0.35/t_{r,\text{true}} = 175 \,\si{\mega\hertz}$, so the required
scope bandwidth is about $3.1$ times the signal bandwidth,
consistent with the working rule that a $3$-to-$5\times$ margin
holds the rise-time over-read to a few percent
(\cref{fig:vol01:ch05:scope-risetime}). A designer who buys a
$200 \,\si{\mega\hertz}$ scope for this signal (barely above the
signal's own $175 \,\si{\mega\hertz}$) measures
$t_{r,\text{scope}} = 0.35/(200 \,\si{\mega\hertz}) = 1.75 \,\si{
\nano\second}$ and reads
$t_{r,\text{meas}} = \sqrt{2^{2} + 1.75^{2}} = 2.66 \,\si{\nano
\second}$, a $33\%$ over-read that would mis-diagnose a healthy
edge as slow. The probe's own rise time enters the same quadrature
sum; a $10\times$ passive probe with its own $0.5 \,\si{\nano
\second}$ rise time adds another term the designer must budget.

The working oscilloscope tier in 2026. Entry-level: Rigol DS1054Z
($50\,\si{\mega\hertz}$ bandwidth, $1\,\text{GS/s}$, $8$-bit,
\$$400$); Siglent SDS1104X-E ($100\,\si{\mega\hertz}$,
$1\,\text{GS/s}$, $8$-bit, \$$600$). Mid-range: Keysight DSOX1204G
($200\,\si{\mega\hertz}$, $2\,\text{GS/s}$, $8$-bit, \$$1500$);
Rohde \& Schwarz RTH1004 portable ($500\,\si{\mega\hertz}$,
$5\,\text{GS/s}$, $10$-bit, \$$7000$). Research-grade: Tektronix
MSO 5 Series ($2\,\si{\giga\hertz}$, $25\,\text{GS/s}$,
$12$-bit, \$$30{,}000$); Keysight UXR-Series ($110\,\si{\giga
\hertz}$, $256\,\text{GS/s}$, $10$-bit, \$$500{,}000$ for the
flagship). Application: hobbyist electronics on Rigol; industrial
embedded development on Keysight DSOX; power-supply switching
analysis on RTH; signal-integrity work on UXR.

Oscilloscopes acquired their working role in electronics
laboratories in the mid-twentieth century (Howard Vollum and
Jack Murdock founded Tektronix in $1946$; Tektronix introduced
the $511$ scope at $10\,\si{\mega\hertz}$ bandwidth in $1949$,
the first triggered scope and the working reference instrument
for two decades). The architecture has evolved through analog
(CRT-based) to digital storage to mixed-signal (with digital
channels) to fully software-defined instruments running on
general-purpose computing platforms. The discipline they
enforce is the same: a measurement is what is on the screen,
not what the operator expected to be on the screen.

\subsection{The micrometer and gauge blocks}

A \engterm{micrometer} measures small linear dimensions through a
calibrated screw thread. A standard machine-shop micrometer has a
$25 \,\si{\milli\meter}$ measuring range with $0.01 \,\si{\milli
\meter}$ resolution; precision micrometers reach $0.001 \,\si
{\milli\meter}$. The instrument has been the standard machine-shop
linear measure since the nineteenth century, and its design has
changed little: a frame, a fixed anvil, a moving spindle driven by
a calibrated screw, and a thimble graduated in fractions of a
turn.

The micrometer's accuracy depends on its calibration (against gauge
blocks traceable to chapter 3's standards), its operator's
technique (parallax, force, thermal expansion of the workpiece),
and its environmental conditions (temperature drift). A reader who
uses a micrometer without temperature compensation in a workshop
that is $5 \,\si{\degreeCelsius}$ different from the calibration
laboratory has introduced a systematic bias.

\paragraph{The zero-setting procedure.} Every micrometer carries an
internal calibration that drifts with handling, thermal cycling, and
mechanical wear, and the working operator zeros the instrument
before each measurement session. For a digital micrometer such as
the Mitutoyo Series 293 \cite{text:ch05t3-mitutoyo-293-datasheet},
the procedure is: clean the anvil and spindle faces with a
solvent-soaked cleaning paper; close the spindle against the anvil
using the ratchet stop (which applies a controlled measurement
force of $\sim 5 \,\si{\newton}$); press the zero/origin button.
For a $0$ to $25 \,\si{\milli\meter}$ micrometer the anvil and
spindle close directly; for a $25$ to $50 \,\si{\milli\meter}$
instrument the operator inserts a calibrated $25 \,\si{\milli
\meter}$ gauge block (or a setting standard supplied with the
instrument) between anvil and spindle and zeros against the gauge
block's certified length. The certificate of the gauge block,
traceable through a calibration chain to a national laboratory
\cite{text:doiron-beers-2009}, is what makes the micrometer's
reading metrologically meaningful.

\paragraph{The Abbe principle.} The micrometer's design places the
measurement line (the centreline of the workpiece between anvil and
spindle faces) parallel to the screw's line of action; the two lines
are not coaxial. Any angular tilt of the spindle relative to the
anvil multiplies the inter-line distance, producing a first-order
error in the measured length. Ernst Abbe stated the working
principle in 1890 \cite{paper:bryan-abbe-1979}: "If errors of
parallax are to be avoided, the measuring system must be placed
coaxially with the line in which displacement is to be measured on
the workpiece." A micrometer with an Abbe offset of $a$ and a
spindle tilt of $\theta$ introduces a length error $a \theta$ in
the small-angle limit. For a standard machine-shop micrometer with
$a \approx 10 \,\si{\milli\meter}$ Abbe offset and a wear-induced
spindle tilt of $\theta \approx 100 \,\si{\micro\radian}$, the
error is $a\theta \approx 1 \,\si{\micro\meter}$. Coordinate
measuring machines, vernier calipers, and dial bore gauges that
violate the Abbe principle pay a corresponding accuracy cost.

\paragraph{Thermal effects.} Steel expands at approximately
$\alpha = 11 \cdot 10^{-6} \,\si{\per\kelvin}$, and dimensional
metrology is conventionally reported at the ISO reference temperature
of $20 \,\degC$ \cite{std:iso-3650}. A $100 \,\si{\milli\meter}$
steel workpiece measured at $25 \,\degC$ rather than $20 \,\degC$
reads $\Delta L = L \alpha \Delta T = 100 \cdot 11 \cdot 10^{-6}
\cdot 5 = 5.5 \,\si{\micro\meter}$ long, an error of $55 \,\si{ppm}$. For a Class A gauge block this is many
times the calibrated tolerance.

\paragraph{Worked precision-dimensional case: the $50 \,\si{\milli
\meter}$ gauge block that grew under the operator's hand.} A
working metrology laboratory uses a $50 \,\si{\milli\meter}$ steel
gauge block (Mitutoyo grade 0, calibrated tolerance $\pm 0.10 \,\si
{\micro\meter}$) to set the zero on a $25$ to $50 \,\si{\milli
\meter}$ micrometer. The operator removes the block from its
storage case, holds it between thumb and forefinger while
demagnetising, places it between the micrometer faces, and reads
$50.0006 \,\si{\milli\meter}$, $0.6 \,\si{\micro\meter}$ long. A
re-measurement after $20 \,\si{\minute}$ in the climate-controlled
soak chamber reads $50.0001 \,\si{\milli\meter}$, within the
calibrated tolerance.

The intermediate reading was not the gauge block's actual length;
it was the gauge block plus the operator's body heat. The contact
patch of thumb and forefinger transferred heat to the block;
\cite{paper:ch05t3-johnson-clarke-2003} reports that a similar
steel block held for $30 \,\si{\second}$ rises by $0.5$ to $1.0
\,\degC$ at the contact patch and shows a length increase of $0.55$
to $1.1 \,\si{\micro\meter}$ across the block. The $0.5 \,\si
{\micro\meter}$ apparent over-length in our case sits inside that
range. The working mitigation is on every metrology-laboratory
checklist: gloves or insulated tongs for gauge-block handling, and
a $20 \,\si{\minute}$ thermal soak in the measurement environment
before reading.

The case is the working lesson of the chapter in miniature. The
micrometer was working. The gauge block was within its calibrated
tolerance. The measurement chain was traceable. The transducer was
the operator's hand, and the calibration chain had not anticipated
it. Sensor selection (section 5.4) named the environment as one
constraint; the operator's body is part of that environment for any
laboratory measurement at the micrometre scale.

\subsection{The scale}

The analytical balance, mass scale, or kitchen scale was the
working mass instrument the reader's project in chapter 1 used.
Modern electronic scales typically use a load cell (a strain-gauge
based force transducer) read through a high-resolution ADC. A
laboratory analytical balance reaches $0.1 \,\si{\milli\gram}$
resolution at $200 \,\si{\gram}$; a household kitchen scale offers
$1 \,\si{\gram}$ at $5 \,\si{\kilogram}$.

The mass measurement is sensitive to local gravity (a scale
calibrated at one latitude reads slightly differently at another),
to air buoyancy on samples of unusual density, to electrostatic
charge on the sample or the pan, and to the level of the
instrument. A precision laboratory balance is more an integrated
measurement system than a single sensor; its operating manual
documents each correction.

\subsection{The microscope}

The \engterm{optical microscope} extends the human eye's resolution
through magnification, traditionally to about $0.2 \,\si{\micro
\meter}$ resolution at visible wavelengths (the diffraction limit).
Modern compound microscopes incorporate digital image sensors,
software stitching, and computational reconstruction.

Beyond optical resolution, SEM, TEM, and AFM trade sample
preparation, vacuum or surface requirements, calibration, and
artefacts for much smaller length scales. Their numerical
resolutions belong with cited microscopy handbook references; here
we only state the ordering. The scanning electron microscope (SEM)
focuses an electron beam onto the specimen and detects backscattered
or secondary electrons; the transmission electron microscope (TEM)
transmits electrons through a thin specimen; atomic force
microscopy (AFM) scans a sharp tip across the specimen surface.
Each is its own measurement chain, with its own calibration
discipline. Volume V (Materials) takes up these instruments in
context.

\subsection{Sensor arrays}

The single sensor is the simplest case. Many modern measurements
use \engterm{sensor arrays} that combine multiple sensing elements
to produce information beyond what any single element can deliver.
Examples include image sensors (millions of photodiodes in a
camera), microphone arrays for direction-of-arrival estimation,
electromyography arrays, accelerometer constellations for
structural health monitoring, and seismometer networks. Sensor
arrays will recur in Volume VII (information) and Volume IX
(systems); for now, the reader should know that the section's
single-sensor framing generalises.

The smartphone is itself a sensor array. A typical smartphone
contains an accelerometer, a gyroscope, a magnetometer, a
barometer, a proximity sensor, an ambient-light sensor, multiple
microphones, multiple cameras (each with its own image sensor),
GPS and other satellite receivers, cellular and Wi-Fi radio
receivers, and an array of capacitive touch sensors on the screen.
The reader's smartphone is, by raw sensor count, one of the most
densely instrumented objects in their daily environment.

\paragraph{Thermistor chains for liquid level.}
A single thermistor at a fixed depth reads the liquid temperature at
that depth. A vertical chain of thermistors on a common cable, with
each element at a stated depth interval, reads the temperature
profile along the chain. The Omega TR-CK series
\cite{text:ch05t3-omega-trck-datasheet} offers $4$ to $32$ elements
of $10 \,\si{\kilo\ohm}$ NTC at $25 \,\degC$ on a $10 \,\si{\meter}$
sheathed cable with $\pm 0.2 \,\degC$ interchangeability. In a
chemical-process tank the discontinuity between liquid (good thermal
contact) and vapour (poor thermal contact) lets the chain report the
liquid level as the boundary between fast-responding and
slow-responding elements after a heater pulse: a working level
sensor built from a thermal-time-constant array rather than a
pressure or ultrasonic transducer. The same array reads the
stratification profile in a thermal-storage tank as a working
sub-product.

\paragraph{Strain-gauge rosettes.}
A single uniaxial strain gauge measures normal strain along one
axis. Principal stresses in a general two-dimensional state cannot
be reconstructed from a single gauge; they require strain
measurements in at least three independent directions. The
strain-gauge rosette places three gauges on a common backing at
known angular offsets (commonly $0$, $45$, $90$ degrees for the
rectangular rosette, or $0$, $60$, $120$ for the delta rosette). The
Micro-Measurements EA-06-250RA-120
\cite{text:ch05t3-micromeas-ea06250ra-datasheet} is a working
example: a $6.35 \,\si{\milli\meter}$ active grid length, $120 \,\si
{\ohm}$ nominal resistance, gauge factor $2.085$, three grids
oriented vertical, $45^{\circ}$ diagonal, and horizontal on a single
flexible backing. The three strain readings
$\varepsilon_{0}, \varepsilon_{45}, \varepsilon_{90}$ reconstruct
the principal strains:
\[
\varepsilon_{1,2} = \frac{\varepsilon_{0} + \varepsilon_{90}}{2}
\pm \frac{1}{\sqrt{2}} \sqrt{(\varepsilon_{0} - \varepsilon_{45})^{2}
+ (\varepsilon_{45} - \varepsilon_{90})^{2}}.
\]
The principal-stress directions follow from the rosette angle and
the principal-strain values; the principal stresses follow from
Hooke's law in plane stress, taken up in Volume V (Materials). The
sensor-engineering point is that the rosette is a working sensor
array whose three elements share a common backing, common
temperature, and common bonding adhesive: shared environment is a
feature here, since temperature compensation cancels in the
difference. Independent failure of one grid is rare; common-mode
failure (rosette debonded as a unit) is the dominant field mode.

\paragraph{Image-sensor demosaicing.}
A modern colour image sensor is an array of millions of monochrome
photodiodes with a colour-filter mosaic deposited on top. The
\engterm{Bayer pattern}, due to Bryce Bayer at Eastman Kodak in
1976, alternates red, green, and blue filters in a $2 \times 2$
repeating tile with two greens per tile (matching the human eye's
peak luminance sensitivity in the green band). A Nikon D850 image
sensor reads $45.7$ megapixels of monochrome photodiode under a
Bayer filter; the camera's image processor \engterm{demosaics} the
raw read by interpolating the missing colour channels at each pixel
location, producing a full-resolution RGB image. The demosaicing
algorithm is itself a calibration function from the sensor's
$45.7\,\text{M}$ monochrome counts to the $45.7\,\text{M}$
RGB triples the operator reads. A different demosaicing algorithm
applied to the same raw read produces a measurably different image;
the choice of algorithm is part of the working sensor-array
calibration that the camera firmware commits to without the operator
seeing the choice.

\paragraph{Cross-talk in multi-channel acquisition.}
A sensor array's elements share a wiring harness, a connector, and
often an ADC. Cross-talk between channels is the regime where a
signal injected on channel $i$ appears, attenuated, on channel
$j \ne i$ at the ADC output. The National Instruments NI 9213
sixteen-channel thermocouple input module
\cite{text:ch05t3-ni-9213-datasheet} specifies channel-to-channel
crosstalk as less than $-100 \,\si{\decibel}$ in the high-resolution
mode, meaning a $1 \,\si{\milli\volt}$ thermocouple signal on
channel $1$ contributes less than $10 \,\si{\nano\volt}$ to the
reading on channel $2$. For a type-K thermocouple at $\sim 40 \,\si
{\micro\volt}/\degC$, this is $\sim 0.25\,\text{m}\degC$ of
apparent reading on the adjacent channel from the first channel's
working signal, working well inside the module's $\pm 0.07 \,\degC$
type-K accuracy. The same module's $250 \,\si{\volt}$
channel-to-bus isolation is the working figure that matters when
one thermocouple is grounded to a $230 \,\si{\volt}$-mains-powered
piece of process equipment and the next channel is grounded to the
acquisition chassis. Cross-talk is a working sensor-array spec, not
a footnote; reading the data sheet's channel-isolation and
channel-cross-talk lines before specifying the module is part of
the sensor-array selection discipline.

\paragraph{Sensor fusion: the complementary filter and a Kalman
preview.} A sensor array often delivers redundant measurements of
the same underlying quantity through different physical principles
with different error structures. A drone's attitude estimate
combines an accelerometer (which reports the gravity vector but is
corrupted by linear acceleration), a gyroscope (which reports the
angular rate with sub-degree short-term accuracy but drifts at
$\sim 5 \,\si{\degree}/\si{\hour}$ from bias), and a magnetometer
(which reports the local magnetic-field vector but is corrupted by
nearby ferrous metal). A working autopilot must produce a single
attitude estimate from these three sensors, each of which lies in a
different way. The working machinery is \engterm{sensor fusion}.

The simplest fusion algorithm is the
\engterm{complementary filter}: a one-line equation that uses the
gyroscope for short-term dynamics and the accelerometer for
long-term reference. For a pitch-angle estimate $\hat{\theta}$, the
recursion is
\[
\hat{\theta}_{k+1} = \alpha \, \big( \hat{\theta}_{k} +
\omega_{k} \Delta t \big) + (1 - \alpha) \, \theta_{\text{acc},k+1},
\]
where $\omega_{k}$ is the gyroscope's angular rate at step $k$,
$\Delta t$ is the sample period, $\theta_{\text{acc}}$ is the
accelerometer-derived pitch (computed from
$\arctan(a_{x} / a_{z})$), and $\alpha \approx 0.98$ for typical
$\Delta t = 5 \,\si{\milli\second}$ tuning. The filter trusts the
integrated gyroscope rate over short intervals (the gyroscope's
short-term noise floor is far below the accelerometer's
linear-acceleration corruption) and trusts the accelerometer over
long intervals (the accelerometer's gravity-vector reading does not
drift, while the gyroscope integrates bias into unbounded angle
error). The $\alpha = 0.98$ tuning corresponds to a high-pass corner
of $\sim 1.6 \,\si{\hertz}$ on the gyroscope path and a low-pass
corner of $\sim 1.6 \,\si{\hertz}$ on the accelerometer path,
chosen so that the gyroscope handles dynamics above the corner and
the accelerometer handles the long-term gravity reference.

The \engterm{Kalman filter} is the working generalisation. The
Honeywell HG1700 \cite{text:ch05t3b-honeywell-hg1700-datasheet}
inertial measurement unit (IMU), at \$$10{,}000$ for the tactical
grade, packs three ring-laser gyroscopes (Honeywell GG1320, drift
$1 \,\si{\degree}/\si{\hour}$) and three accelerometers
(QA-2000 family, bias stability $25 \,\mu g$) onto a common
chassis with synchronised digital outputs at $600 \,\si{\hertz}$. A
working navigation system fuses the HG1700's body-frame angular
rates and accelerations with intermittent GPS position fixes (at
$1$-to-$10 \,\si{\hertz}$, $\sim 3 \,\si{\meter}$ accuracy) through
a Kalman filter that maintains an estimate of position, velocity,
attitude, gyro bias, and accelerometer bias as a $15$-state vector,
updated at each sample by the IMU's high-rate measurements and
corrected at each GPS fix by the position observation. Between GPS
fixes the IMU integrates inertial dynamics with the Kalman-tracked
biases removed; at each fix the Kalman update redistributes the
position error across the $15$ states, tightening the bias estimates
and resetting the integrated drift. The working result: a navigation
solution at $600 \,\si{\hertz}$ with position accuracy held to
$\sim 5 \,\si{\meter}$ during $30 \,\si{\second}$ GPS outages,
beyond what any single sensor could deliver. The mathematical
machinery of the Kalman filter (state-transition model, measurement
model, Kalman gain, covariance propagation) waits for Volume VI
chapter 4 on time-frequency duality (where the state-space
formulation lives) and Volume VII chapter 15 on state-space
estimation (where the optimality, observability, and tuning
discipline live).

The sensor-fusion idea generalises: every working sensor array that
combines elements with different error structures is a candidate
for a fusion algorithm that extracts more than the per-element
information. A car's stability-control system fuses wheel-speed
sensors, yaw-rate gyroscope, steering-wheel angle, and lateral
accelerometer to estimate vehicle slip angle; a smartphone fuses
GPS, accelerometers, gyroscopes, magnetometer, barometer, and
Wi-Fi RSSI to estimate position at $\sim 5 \,\si{\meter}$
horizontal accuracy indoors where the GPS fix alone is unreliable;
a power-plant condition monitoring system fuses vibration,
temperature, oil-debris, and electrical-current sensors to estimate
bearing remaining-useful-life. The sensor-engineering point: the
single-sensor framing of section 5.2 is the starting case; the
sensor-fusion lane is the working frontier where the chapter's
material connects forward to Volumes VI, VII, and IX.

The Kalman family extends along two working axes when the
linear-Gaussian assumptions break down. The
\engterm{extended Kalman filter (EKF)} linearises the
state-transition and measurement models at each step around the
current state estimate, trading the optimality of the linear case
for tractability on nonlinear navigation, target tracking, and
robotic-arm kinematics. The \engterm{unscented Kalman filter (UKF)}
propagates a set of deterministic sample points (the sigma points)
through the nonlinear models and reconstructs mean and covariance
from the propagated samples, avoiding the Jacobian computation the
EKF requires; the working trade-off is computational cost against
linearisation error. The \engterm{particle filter} drops the
Gaussian assumption entirely and represents the state as a sample
of weighted Monte Carlo draws, propagated through the nonlinear
models and resampled at each measurement update; the working choice
when the state distribution is multimodal (a robot uncertain between
two corridors, a target track that may correspond to either of two
nearby blips). Each extension carries a working complexity cost:
the EKF runs at a few times the linear-Kalman cost, the UKF at an
order of magnitude more, the particle filter at the cost of
thousands of particles per update. The working selection follows
the working measurement: linear-Gaussian for an HG1700-grade IMU
between GPS fixes, EKF for a Tesla Autopilot's vehicle dynamics
fused with map data, UKF for a quadrotor's attitude estimate during
aggressive manoeuvres, particle filter for a Mars-rover SLAM that
must consider multiple terrain hypotheses simultaneously.

The fault-detection branch of sensor fusion runs alongside the
state-estimation branch. A \engterm{residual} is the difference
between a sensor's measurement and the fused estimator's prediction
of that measurement; a residual much larger than its predicted
covariance flags a sensor as anomalous and triggers the working
fault-isolation procedure. A working triplex-redundant air-data
system (three pitot tubes, three angle-of-attack vanes, three total-
temperature probes) runs a chi-squared test on the residuals at
every sample, isolates a divergent sensor, votes out the divergent
sample, and continues operation on the working two-of-three majority
until the maintenance interval. The same residual machinery powers
the predictive-maintenance lane: a working bearing's vibration
spectrum diverges from its history at the working failure onset,
the residual against the working baseline flags the working anomaly,
and the working maintenance team intervenes before the bearing
fails. The Kalman residual is, in this working role, a working
discriminator that the sensor array could not produce element by
element. The same discriminator returns in Volume IX chapter 8 on
fault-detection-and-isolation for working flight-critical systems.

\subsection{The spectrum analyser}

A \engterm{spectrum analyser} displays signal amplitude versus
frequency. The working benchtop instrument is the swept-tuned
super-heterodyne (the Keysight N9020B MXA is representative); a
modern alternative is the real-time spectrum analyser, which uses
fast Fourier transform on a continuously sampled IF and trades
single-sweep speed for gap-free time coverage. Three specifications
distinguish working instruments.

\engterm{Resolution bandwidth (RBW).} The width of the IF filter
through which the swept-tuned local oscillator passes the signal of
interest. A $1 \,\si{\hertz}$ RBW resolves two tones separated by a
few hertz; a $1 \,\si{\mega\hertz}$ RBW does not. The sweep time
through a given span scales as the inverse square of RBW for the
single-pole IF filter (longer sweep at narrower RBW). The Keysight
N9020B offers RBW from $1 \,\si{\hertz}$ to $8 \,\si{\mega\hertz}$
\cite{text:ch05t3-keysight-n9020b-datasheet}.

\engterm{Video bandwidth (VBW).} The width of a post-detector
low-pass filter that averages the noise floor. Setting VBW below
RBW smooths the displayed trace and reduces the variance of the
displayed noise floor (by approximately $\sqrt{\text{VBW}/\text{RBW}}$
in amplitude). VBW does not change the absolute noise floor; it
reduces the apparent jitter of the noise floor on the display, the
working variable when reading a small signal off a noisy trace.

\engterm{Displayed average noise level (DANL).} The noise floor of
the analyser itself, normalised to $1 \,\si{\hertz}$ bandwidth. The
working figure for the Keysight N9020B with the preamplifier on at
$2 \,\si{\giga\hertz}$ is approximately $-166 \,\text{dBm/Hz}$
\cite{text:ch05t3-keysight-n9020b-datasheet}. The displayed noise
floor at a given RBW is $\text{DANL} + 10 \log_{10}(\text{RBW}/\si
{\hertz})$. A $1 \,\si{\kilo\hertz}$ RBW raises the displayed floor
to $-136 \,\text{dBm}$; a $100 \,\si{\kilo\hertz}$ RBW raises it to
$-116 \,\text{dBm}$. A signal $20 \,\si{\decibel}$ below the
displayed floor is undetectable; the working operator reduces RBW
to expose a buried tone, at the cost of sweep time.

\engterm{Dynamic range.} The ratio of the largest signal the
analyser can handle without distortion (set by the third-order
intercept point, TOI) to the noise floor. The Keysight N9020B
specifies TOI $\approx +22 \,\text{dBm}$ at $2 \,\si{\giga\hertz}$
and DANL $\approx -166 \,\text{dBm/Hz}$, for a spurious-free dynamic
range of approximately $(2/3)(\text{TOI} - \text{DANL}) \approx 125
\,\si{\decibel}$ in $1 \,\si{\hertz}$ RBW.

The working spectrum-analyser tier in $2026$: entry-level Siglent
SSA3015X ($1.5 \,\si{\giga\hertz}$, $\$1{,}500$);
mid-range Rigol RSA5065N ($6.5 \,\si{\giga\hertz}$ with real-time
mode, $\$10{,}000$); benchtop Keysight N9020B MXA ($50
\,\si{\giga\hertz}$, $\$40{,}000$); research-grade Rohde \&
Schwarz FSV40 ($40 \,\si{\giga\hertz}$ with high-end RF performance,
$\$100{,}000$ and up). Application: EMC pre-compliance scans
on Siglent; software-defined-radio spurious-emission tests on Rigol;
production-test radio-frequency measurements on Keysight;
phase-noise and millimetre-wave characterisation on R\&S.

\paragraph{Reading a working trace.} A spectrum analyser display
shows frequency on the horizontal axis (centre frequency
$\pm$ span/$2$) and amplitude in dBm on the vertical axis. The
working operator places markers (Keysight calls them M1, M2,
$\ldots$) at signals of interest; the marker readout reports
frequency and amplitude to the analyser's resolution. A peak
marker tracks the highest signal in the trace; a delta marker
reports frequency and amplitude relative to a reference marker, the
working way to measure harmonic spacing and modulation sidebands.
A signal of interest on a noisy trace is identified by lowering
RBW until the signal stands above the displayed noise floor by at
least $\sim 10 \,\si{\decibel}$, then reading the marker. A signal
that does not appear above DANL $+ 10 \log_{10}(\text{RBW}/\si
{\hertz})$ at the narrowest available RBW is below the analyser's
sensitivity for that frequency. The working operator at this point
either uses a preamplifier (lowering DANL by the preamplifier
gain, at the cost of compression for any in-band strong signal) or
accepts that the analyser is not the right instrument for that
signal.

\subsection{The digital lock-in amplifier}

A \engterm{lock-in amplifier} recovers a small signal at a known
frequency from a noise background many orders of magnitude larger.
The working principle is phase-sensitive detection: the noisy input
is multiplied by a reference at the signal's frequency, producing a
DC term proportional to the in-phase component of the signal of
interest plus AC terms at the sum and difference frequencies of
every noise component; the AC terms are then removed by a low-pass
filter at a long time constant, leaving the DC term. A second
mixer at $90^{\circ}$ recovers the quadrature component, and the
two together give magnitude and phase. The lock-in was first
described in working form by Walter Michels and N. L. Curtis in
$1941$ as a pentode-based phase-sensitive detector
\cite{paper:ch05t3-michels-curtis-1941}; the digital lock-in,
in which the mixers and low-pass filters live in DSP, became the
working instrument in the $1980$s and $1990$s. The Stanford
Research Systems SR860 ($\$9{,}500$,
\cite{text:ch05t3-srs-sr860-datasheet}) is representative.

The working specifications. The SR860 covers reference frequencies
$4 \,\si{\milli\hertz}$ to $500 \,\si{\kilo\hertz}$, with input
equivalent noise $2.5 \,\si{\nano\volt}/\sqrt{\si{\hertz}}$ at $1
\,\si{\kilo\hertz}$ and a dynamic reserve of $120 \,\si{\decibel}$
(the ratio of the largest interfering signal the input can tolerate
without overload to the smallest signal of interest the instrument
can recover). Time constants are selectable from $1 \,\si{\micro
\second}$ to $30{,}000 \,\si{\second}$. The working trade-off
\cite{paper:ch05t3-meade-1983}: a long time constant gives narrower
equivalent noise bandwidth and a lower noise floor, at the cost of
slower response to changes in the signal of interest. For a single-
pole low-pass with time constant $\tau$, the equivalent noise
bandwidth is $1/(4\tau)$.

\paragraph{Worked example: recovering a $0.1 \,\si{\hertz}$
modulation buried in $1/f$ noise.}
An optical transmittance experiment chops a probe beam at
$f_{\text{ref}} = 137 \,\si{\hertz}$ (chosen away from any line
harmonic, far from the $1/f$ knee of the photodiode amplifier, and
not a multiple of $50$ or $60 \,\si{\hertz}$). The detected
photocurrent contains a $137 \,\si{\hertz}$ tone of amplitude
$50 \,\si{\pico\ampere}$, riding on a background of
$\sim 1 \,\si{\micro\ampere}$ of $1/f$ noise integrated from
$0.01$ to $10 \,\si{\hertz}$. The signal of interest is the slow
$0.1 \,\si{\hertz}$ modulation of the $137 \,\si{\hertz}$
amplitude as a chemical concentration in the optical path drifts.

The photodiode current passes through a $10^{8} \,\si{\ohm}$
transimpedance amplifier, yielding $\sim 5 \,\si{\milli\volt}$ at
$137 \,\si{\hertz}$ riding on $\sim 100 \,\si{\milli\volt}$ of
$1/f$ noise at the input of the lock-in. The lock-in's reference
input is tied to the chopper's TTL output at $137 \,\si{\hertz}$,
so the working frequency is locked. The mixer multiplies the input
by $\cos(2\pi f_{\text{ref}} t)$; the $1/f$ noise, almost entirely
below $10 \,\si{\hertz}$, multiplied by the $137 \,\si{\hertz}$
reference produces sidebands at $\sim 127$ to $147 \,\si{\hertz}$
that the low-pass filter must remove. The $0.1 \,\si{\hertz}$
modulation of the signal of interest sits at $f_{\text{ref}} \pm 0.1
\,\si{\hertz}$ at the mixer input and at $\pm 0.1 \,\si{\hertz}$ at
the mixer output. To recover the modulation we need a time constant
that passes $0.1 \,\si{\hertz}$ and blocks the $1/f$ sidebands. A
$\tau = 0.3 \,\si{\second}$ time constant gives equivalent noise
bandwidth $\sim 0.8 \,\si{\hertz}$, passing the $0.1 \,\si{\hertz}$
modulation with $\sim 2 \,\si{\decibel}$ attenuation and rejecting
the $1/f$ sidebands by $\sim 30 \,\si{\decibel}$ at $\sim 5 \,\si
{\hertz}$. A longer $\tau = 3 \,\si{\second}$ would lower the noise
floor further but would also attenuate the $0.1 \,\si{\hertz}$
modulation by a factor of $\sim 8$ in amplitude, defeating the
working measurement. The working choice sits in the middle.

The lock-in amplifier turns a chemical-concentration measurement
that would be invisible in a direct read into a working signal with
working SNR. The working price is that the experiment must inject a
known modulation at a known frequency in the first place; the
lock-in cannot recover a signal whose frequency is unknown to it.
For an absorbed-photon experiment the modulation is the
beam-chopper; for a magnetic-resonance experiment it is the
modulation coil; for a phase-locked-loop control system it is the
oscillator's reference. Every working lock-in measurement carries
the same architectural commitment: pick a frequency, modulate the
quantity of interest at that frequency, demodulate, low-pass.
Volume VI chapter 4 on time-frequency duality returns to this
narrow-band recovery in the language of windowed Fourier
transforms; the lock-in is the working hardware realisation of the
mathematical machinery that volume develops.

\subsection{The digital boxcar averager}

A \engterm{boxcar averager} recovers a repetitive pulsed signal by
integrating the input over a gated window timed to the pulse, then
averaging the gate-window integrals across many pulses. Where the
lock-in is the working instrument for continuous narrow-band
signals, the boxcar is the working instrument for repetitive
transients whose duty cycle is small, whose timing is known, and
whose between-pulse interval carries no signal of interest. The
Stanford Research Systems SR250 gated integrator and boxcar
averager \cite{text:ch05t3b-srs-sr250-datasheet} is the working
benchtop instrument: \$$5{,}000$ at the time of writing, gate
widths from $2 \,\si{\nano\second}$ to $15 \,\si{\micro\second}$,
trigger-to-gate jitter $< 50 \,\si{\pico\second}$, sample rates to
$20 \,\si{\kilo\hertz}$, and an internal exponential moving-average
filter that delivers $N$-pulse averaging with selectable equivalent
time constants from $1$ to $10{,}000$ pulses.

The working measurement. A laser-pulse experiment fires a pulse of
$\sim 5 \,\si{\nano\second}$ duration at $10 \,\si{\kilo\hertz}$
repetition rate; a photomultiplier views the laser-induced
fluorescence; the working signal is a fluorescence transient
$\sim 50 \,\si{\nano\second}$ wide whose peak is buried in
photomultiplier shot noise. The boxcar gates on the trigger from
the laser's Q-switch driver, opens an integration window
$10 \,\si{\nano\second}$ before the fluorescence peak, closes the
window $60 \,\si{\nano\second}$ later, and integrates the
photomultiplier current over that window. Between pulses the
photomultiplier sees only ambient and dark current; the boxcar
ignores the between-pulse interval entirely. After $N$ pulses the
averager has summed $N$ gate-window integrals. The signal-to-noise
ratio scales as $\sqrt{N}$: $10{,}000$ pulses delivers a
$100\times$ SNR improvement over a single pulse, and at the
$10 \,\si{\kilo\hertz}$ repetition rate that gain is bought in
$1 \,\si{\second}$ of measurement time.

The duty-cycle gain is the working figure. A
$50 \,\si{\nano\second}$ gate at $10 \,\si{\kilo\hertz}$ repetition
rate has an on-time fraction of $5 \times 10^{-4}$; a
DC-coupled amplifier looking at the same photomultiplier sees the
between-pulse noise at all times, so its noise-equivalent-bandwidth
is the full measurement bandwidth. The boxcar's effective bandwidth
is the gate-width bandwidth times the duty cycle, a reduction of
$2000\times$ in noise power at fixed signal power, equivalent to
$33 \,\si{\decibel}$ of working SNR gain before any averaging is
applied. The averager's $N$-pulse gain stacks on top of that.

\paragraph{Lock-in or boxcar.} The working choice between lock-in
and boxcar follows the signal's working duty cycle. A signal that
fills the measurement interval at a known frequency (continuous-
wave chopped optical absorption, sinusoidal modulation of a magnetic
moment, AC excitation of a strain bridge) is a lock-in case: the
modulation lives in a narrow band, and the lock-in's phase-sensitive
detection recovers the in-band amplitude at the modulation frequency.
A signal that lives in a small fraction of the measurement interval
at a known repetition rate (laser-induced fluorescence, time-of-
flight mass spectrometry, pulsed-NMR free-induction decay,
single-photon counting from a Geiger-mode avalanche photodiode) is
a boxcar case: the signal lives in the gate window, and the boxcar
discards the between-gate interval where the signal of interest is
absent. The Stanford Research catalogue makes the architectural
point explicit by offering both the SR860 lock-in and the SR250
boxcar from the same maker; the working engineer picks the
instrument whose architecture matches the signal's working duty
cycle, not the more familiar instrument.

\subsection{The instrument arc: from artisanal calibration to the
datasheet}

The modern engineer reads a datasheet, picks an instrument, and
expects the manufacturer's stated accuracy to hold. That
expectation rests on a century-long working tradition. We sketch
the arc.

\paragraph{1888: the cadmium-mercury Weston cell.} Edward Weston
founded the Weston Electrical Instrument Company in Newark, New
Jersey in $1888$, and patented the cadmium-mercury voltaic cell
that became the working voltage standard for laboratories worldwide
\cite{text:ch05t3-fink-electronics-1989}. The Weston cell at
$20 \,\degC$ produces $1.01865 \,\si{\volt}$ with
$\sim 5 \,\si{ppm}$ year-to-year stability when
unloaded, and served as the legal volt in the United States from
$1908$ until the 1972 transition to the Josephson volt. Weston's
ammeters, voltmeters, and galvanometers, with their distinctive
two-sided pivot bearing and the moving-coil mechanism that became
the working laboratory standard, defined the look and feel of the
working laboratory instrument for half a century.

\paragraph{1892: General Electric.} The merger of Edison General
Electric and Thomson-Houston Electric Company in $1892$ created
General Electric, whose instrument lineage flowed through the GE
Meter Department and the later GE Industrial Systems instrument
group. GE's contribution to the working instrument was the
production of switchboard instruments, line-frequency meters, and
the early electronic measurement instruments at scale: the
working instrument moved from the bench-top artisan craft of
Weston's New Jersey shop to the production line.

\paragraph{1939: the HP 200A.} William Hewlett and David Packard,
recent Stanford graduates working out of Packard's garage in Palo
Alto, built the HP 200A audio oscillator in $1939$ as Hewlett's
master's-thesis project: a variable-frequency Wien-bridge RC
oscillator with an incandescent-bulb amplitude-stabilisation loop,
producing low-distortion sine waves from $20 \,\si{\hertz}$ to
$20 \,\si{\kilo\hertz}$ for $\$71.50$ a unit
\cite{web:ch05t3-hp-200a-history}. Walt Disney Studios bought eight
HP 200B oscillators to test the surround-sound system for the
$1940$ film \emph{Fantasia}, the first production sale and the
beginning of the Hewlett-Packard test-and-measurement business
\cite{web:ch05t3-agilent-keysight-history}.

\paragraph{1949 and 1972: the Tektronix oscilloscope.} Howard
Vollum and Jack Murdock founded Tektronix in Portland, Oregon, in
$1946$, and introduced the model $511$ in $1949$ as the first
triggered oscilloscope, displacing the analog free-running scope
as the working laboratory instrument. The Tektronix $7000$-series
modular mainframes (introduced $1969$ through $1972$,
\cite{web:ch05t3-tek-7000-history}) carried plug-in vertical and
horizontal amplifiers in a common chassis: a single $7704$
mainframe with a $7A22$ differential amplifier plug-in and a $7B92A$
dual time-base plug-in covered the working bench-top oscilloscope
need for two decades, until the digital-storage transition of the
$1980$s rendered the modular architecture commercially superfluous.

\paragraph{1970s: the GPIB era.} The Hewlett-Packard Instrument Bus
(HP-IB), standardised as IEEE 488 in $1975$
\cite{std:ch05t3-ieee-488-1975}, allowed a working laboratory's
instruments to communicate with a controlling computer over a
shared parallel bus. The HP 3478A $5.5$-digit bench multimeter
\cite{text:ch05t3-keysight-3478a-manual}, the HP 3325A
function generator, and the HP $54100$-series oscilloscopes
became the working bench instruments of the $1980$s automation
laboratory: a controller (an HP $9000$ workstation or an IBM PC)
programmed the instruments, captured the measurements, and stored
the data. The working laboratory data sheet, in this era, was a
typewriter-set document with a hand-drawn block diagram on the
inside cover and a $5.5$-digit specification table at the back.

\paragraph{1999 and 2014: HP, Agilent, Keysight.} Hewlett-Packard
spun off its test-and-measurement business to Agilent Technologies
in $1999$, retaining the computer and printer businesses under the
HP name. Agilent in turn spun off its electronic-measurement
business to Keysight Technologies in $2014$, retaining the
chemical-analysis and life-sciences businesses under the Agilent
name \cite{web:ch05t3-agilent-keysight-history}. The working
benchtop instruments the reader will see on a modern engineering
laboratory bench: HP $3458A$ multimeter ($1989$), Agilent
$34401A$ multimeter ($1991$), Keysight $34465A$ multimeter
($2013$). Three names, one continuous engineering organisation,
one working datasheet tradition.

\paragraph{Why the datasheet reads the way it does.} The modern
sensor or instrument datasheet has a stable structure: the front
page names the part and its working role; the specifications page
lists accuracy as a unit of error plus a fraction of full-scale of
error (a working convention that captures both the offset and the
gain-error of a typical instrument); the operating-ranges page
states the conditions of the specification (temperature, humidity,
warm-up time, calibration interval); the small-print page covers
the conditions under which the specification does not hold. The
working engineer's habit, of reading the operating-ranges page
before reading the specifications page, is the discipline that
carries forward from the Weston Electrical Instrument Company's
1888 datasheet for the cadmium-mercury cell to the Keysight 34470A
datasheet at the time of the present chapter. The working datasheet
is itself a calibration chain: the unit of error has its source in
the manufacturer's house standard, traceable through the national
laboratory to the SI; the fraction of full-scale of error has its
source in the manufacturer's design conventions. Reading both is
part of the working sensor-selection discipline section 5.4
codified.

\subsection{The instrumentation amplifier}

Between the sensor and the ADC sits the signal-conditioning
amplifier. A thermocouple delivers $\sim 40 \,\si{\micro\volt}/
\degC$; a $350 \,\si{\ohm}$ strain-gauge bridge at
$2 \,\si{\milli\volt}/\si{\volt}$ excitation delivers
$10 \,\si{\milli\volt}$ at full scale on a $5 \,\si{\volt}$ supply;
a thermopile from a $15 \,\si{\watt\per\meter\squared}$ heat-flux
probe delivers $\sim 30 \,\si{\micro\volt}$ per working measurement.
None of these signals can be read directly by an ADC: the working
ADC expects a $\pm 5 \,\si{\volt}$ or $\pm 10 \,\si{\volt}$ input
range, and a $30 \,\si{\micro\volt}$ signal at the ADC input is
below the LSB of any architecture short of the metrology DMM. The
signal-conditioning chain amplifies the sensor's differential output
by a factor of $100$ to $1000$, rejecting the common-mode component
(the ground potential that lifts both inputs together) that would
otherwise saturate the amplifier or appear as a signal-dependent
error at the ADC. The working instrument is the
\engterm{instrumentation amplifier}.

\paragraph{The three-op-amp instrumentation amplifier.} The classical
architecture, due to a 1960s refinement of the differential
amplifier, places three operational amplifiers in a configuration
that delivers high input impedance, high common-mode rejection ratio
(CMRR), and a gain set by a single external resistor. The input
stage uses two operational amplifiers in non-inverting configuration,
each with unity gain to its own input and a gain to the other input
set by the gain-setting resistor $R_{G}$; the output of the input
stage is a differential signal $(V_{+} - V_{-})(1 + 2R/R_{G})$ where
$R$ is the fixed feedback resistor and $R_{G}$ the external
gain-setting resistor. The second stage is a difference amplifier
that converts the differential output of the input stage to a
single-ended output referenced to the amplifier's reference pin. CMRR depends on
the matching of the second-stage resistors; the integrated-circuit
instrumentation amplifier laser-trims those resistors during
manufacturing to deliver $80$ to $130 \,\si{\decibel}$ CMRR at the
part level, well beyond what a discrete construction can deliver.

\paragraph{The working parts in 2026.} The Analog Devices AD620
\cite{text:ch05t3b-adi-ad620-datasheet} is the workhorse
single-channel instrumentation amplifier at
\$$5$ per part: $80 \,\si{\decibel}$ minimum CMRR at unity gain,
$110 \,\si{\decibel}$ at $G = 100$, $50 \,\si{\micro\volt}$ maximum
input offset, $13 \,\si{\nano\volt}/\sqrt{\si{\hertz}}$ input
voltage noise, and $120 \,\si{\kilo\hertz}$ bandwidth at $G = 100$.
The Analog Devices AD8221
\cite{text:ch05t3b-adi-ad8221-datasheet} (\$$8$) replaces the
AD620 in precision applications: $90 \,\si{\decibel}$
minimum CMRR at unity gain, $120 \,\si{\decibel}$ at $G = 100$,
$25 \,\si{\micro\volt}$ maximum input offset,
$8 \,\si{\nano\volt}/\sqrt{\si{\hertz}}$ input voltage noise. The
Analog Devices AD8429 \cite{text:ch05t3b-adi-ad8429-datasheet}
(\$$12$) is the ultra-low-noise instrumentation amplifier
for strain-gauge and thermocouple readouts:
$110 \,\si{\decibel}$ minimum CMRR at unity gain, $130
\,\si{\decibel}$ at $G = 100$, $1 \,\si{\nano\volt}/\sqrt{\si{
\hertz}}$ input voltage noise, and $1.2 \,\si{\mega\hertz}$
bandwidth at $G = 100$. The ladder from AD620 to AD8221 to
AD8429 is a price-versus-CMRR ladder: at $G = 100$, the
amplified-signal-to-common-mode-interference ratio is
$80 \,\si{\decibel}$ (factor $10^{4}$) on the AD620,
$100 \,\si{\decibel}$ (factor $10^{5}$) on the AD8221,
and $110 \,\si{\decibel}$ (factor $3 \times 10^{5}$) on the AD8429.
For a strain-gauge bridge on a machine grounded to a
$230 \,\si{\volt}$-mains-powered piece of equipment, a
$1 \,\si{\volt}$ common-mode noise injected by power-line coupling
appears at the amplifier output as $100 \,\si{\micro\volt}$ (AD620),
$10 \,\si{\micro\volt}$ (AD8221), or $3 \,\si{\micro\volt}$
(AD8429). The choice between the three is the choice between \$$5$
and a $100 \,\si{\micro\volt}$ noise floor, \$$8$ and a
$10 \,\si{\micro\volt}$ noise floor, or \$$12$ and a
$3 \,\si{\micro\volt}$ noise floor.

\paragraph{The chopper-stabilised instrumentation amplifier.} A
class of signal-conditioning applications demands an
input offset below $1 \,\si{\micro\volt}$ and an input-offset drift
below $0.01 \,\si{\micro\volt}/\degC$: thermocouple readouts
at the cold-junction reference, thermopile readouts at the
heat-flux probe, strain-gauge bridges at the zero offset of the
bonded gauge. The chopper-stabilised amplifier delivers those
numbers by modulating the input at a chopper frequency (typically
$10$ to $200 \,\si{\kilo\hertz}$), amplifying the modulated signal
in an AC-coupled stage that has no DC offset, then demodulating
back to DC. The input offset that survives the chopper-demodulator
cycle is the residual after the chopper's switching mismatch;
integrated-circuit chopper-stabilised amplifiers reduce that
residual to $\sim 0.5 \,\si{\micro\volt}$.

The Linear Technology (now Analog Devices) LTC2050
\cite{text:ch05t3b-lt-ltc2050-datasheet} (\$$3$) is the
chopper-stabilised operational amplifier of the working bench:
$0.5 \,\si{\micro\volt}$ maximum input offset,
$0.01 \,\si{\micro\volt}/\degC$ input-offset drift, and
$1.5 \,\si{\micro\volt}/\sqrt{\si{\hertz}}$ input voltage noise at
$1 \,\si{\hertz}$. The architecture's price is a higher input noise
at low frequency than a continuous amplifier, traded for the
offset-and-drift performance. The working bench uses the LTC2050 as
a precision-reference buffer, a strain-gauge single-ended amplifier,
and a thermopile readout where absolute-DC accuracy matters more
than low-noise performance. For a configuration that needs both the
low-offset and the differential-input architecture, the
chopper-stabilised instrumentation amplifier (Analog Devices AD8237
at \$$3$ per part, $5 \,\si{\micro\volt}$ maximum offset,
$0.04 \,\si{\micro\volt}/\degC$ drift) is the single-chip solution.

\paragraph{Worked example: a $350 \,\si{\ohm}$ strain-gauge bridge.}
A machine-frame instrumentation reads strain on a
$350 \,\si{\ohm}$ strain-gauge bridge at $2 \,\si{\milli\volt}/
\si{\volt}$ sensitivity, excited at $5 \,\si{\volt}$ DC. The
full-scale differential output is $10 \,\si{\milli\volt}$; the
sensitivity is $\sim 1 \,\si{\micro\volt}$ per
$\sim 1 \,\si{\micro\strain}$ (the strain unit, a dimensionless
number $10^{-6}$). The downstream ADC expects a $\pm 5
\,\si{\volt}$ input; the required gain is $G = 500$.

A $1 \,\si{\volt}$ common-mode interference on the bridge (coupled
from a $230 \,\si{\volt}$ mains-powered machine frame) appears at
the amplifier output as the common-mode interference times the gain
divided by the CMRR. At $G = 500$, the AD8429's CMRR reaches
$\sim 134 \,\si{\decibel}$ (extrapolated from the $130 \,\si{
\decibel}$ value at $G = 100$ and the $20 \,\si{\decibel}$/decade
increase the datasheet specifies into the part's roll-off region).
The common-mode-induced output: $1 \,\si{\volt} \times 500 /
10^{134/20} = 500 \,\si{\volt} / 5 \times 10^{6} \approx
100 \,\si{\micro\volt}$, equivalent to $\sim 1 \,\si{\micro
\strain}$ of apparent strain at the input. By comparison, the AD620
at $G = 500$ delivers $\sim 100 \,\si{\decibel}$ CMRR; the same
$1 \,\si{\volt}$ common-mode gives an output error of
$1 \,\si{\volt} \times 500 / 10^{100/20} = 5 \,\si{\milli\volt}$,
equivalent to $\sim 50 \,\si{\micro\strain}$, two decades worse than
the AD8429. For a working machine-strain measurement at
$1 \,\si{\micro\strain}$ resolution the AD8429 is the working
choice; for one at $50 \,\si{\micro\strain}$ resolution the AD620
suffices and saves \$$7$. The choice of amplifier sets the noise
floor of the strain measurement; the bridge's sensitivity is the
denominator, not the numerator, of the working uncertainty.

\subsection{The analog-to-digital converter}

Every modern sensor's signal terminates in an analog-to-digital
converter (ADC). The architecture of that ADC determines the
working trade-off between resolution, conversion speed, power
consumption, and cost. Five working architectures cover the
practical sensor-readout landscape.

\engterm{Successive approximation register (SAR)} converters use a
binary-search algorithm: a comparator and digital-to-analog
converter (DAC) iteratively divide the input range until the DAC
output matches the input within $1$ LSB. Conversion time is $N$
clock cycles for $N$ bits. SAR is the workhorse for $8$ to $18$
bits at sample rates up to a few megasamples per second. Working
parts: Analog Devices AD7124-8 ($24$-bit at $19.2\,\text{kS/s}$,
$\$5$); Texas Instruments ADS1256 ($24$-bit
at $30\,\text{kS/s}$, $\$15$); Microchip MCP3008
($10$-bit at $200\,\text{kS/s}$, $\$2$); Analog
Devices AD7980 ($16$-bit at $1\,\text{MS/s}$,
$\$11$). Power consumption: $\sim 1\,\si{\milli\watt}$ per
megasample-per-second for $16$-bit parts.

\engterm{Delta-sigma ($\Delta\Sigma$, sigma-delta)} converters
use oversampling and noise shaping: a low-resolution modulator
(typically $1$-bit) samples at $64$ to $1024$ times the Nyquist
rate, with feedback that shifts quantisation noise out of the
signal band; a downstream digital decimation filter removes the
high-frequency noise and produces the working output at the
output sample rate. The architecture trades sample rate for
resolution: $24$-bit audio ADCs (Asahi Kasei AK5572 at $\$8$
delivering $114\,\si{\decibel}$ dynamic range) are
$\Delta\Sigma$; $32$-bit precision instrument ADCs (Analog
Devices AD7193 at $\$15$, $24$-bit effective for industrial
weighing applications) are $\Delta\Sigma$. The delta-sigma
architecture provides the working precision for the chapter's
strain-gauge bridge readout in the next section. Drawback:
inherent latency from the decimation filter (typically $20$-$100$
samples at the modulator rate) makes $\Delta\Sigma$ unsuitable
for control-loop applications that need single-sample-period
response.

\engterm{Pipeline} converters divide the bits into stages: each
stage samples the input, quantises to a few bits, subtracts the
DAC reconstruction, amplifies the residue, and passes to the
next stage. A $14$-bit converter using $7$ stages with $2$ bits
plus $1$ redundant bit per stage achieves a sample rate equal to
the per-stage rate (typically $100$-$500\,\text{MS/s}$).
Working parts: Analog Devices AD9230 ($12$-bit at
$250\,\text{MS/s}$, $\$80$); Texas Instruments
ADS5400 ($12$-bit at $1\,\text{GS/s}$, $\$200$).
Pipeline is the working choice for software-defined radio,
medical-imaging instrumentation, and digital storage
oscilloscopes from $50$ to $1000\,\text{MS/s}$.

\engterm{Flash} converters quantise the entire input range in
one clock cycle through $2^{N} - 1$ comparators in parallel; the
thermometer-coded comparator outputs are encoded to binary.
Conversion time: one clock cycle, the highest of any
architecture. Resolution is limited to $6$-$10$ bits because the
comparator count scales as $2^{N}$. Working parts: Maxim
MAX109 ($8$-bit at $2.2\,\text{GS/s}$, $\$400$);
Analog Devices HMCAD1511 ($8$-bit at $1\,\text{GS/s}$, $\$50$).
Flash is the architecture for $50\,\text{GS/s}$ scope front-ends,
where the per-sample-period requirement excludes pipeline or SAR.

\engterm{Voltage-to-frequency} converters output a pulse train
whose frequency is proportional to the input voltage; a downstream
counter measures the frequency over a gate time. The working
metrology DMM Keysight 3458A uses a multi-slope voltage-to-
frequency architecture (the "Wakatsuki integrator", U.S. Patent
$4{,}556{,}867$) with $28$-bit resolution at $1\,\si{\hertz}$
sample rate, the precision foundation of the instrument's
$0.1\,\si{ppm}$ absolute accuracy. The voltage-to-frequency
architecture is the working choice when absolute accuracy at low
sample rate is the application's requirement, traded against the
poorer noise performance of the parallel-conversion
architectures.

Architectural choice. For a working temperature reading at
$10\,\text{S/s}$, $\Delta\Sigma$ at $24$-bit
delivers the resolution. For a vibration capture at $50\,\text{kS/s}$,
SAR at $16$-bit suffices. For high-speed transient capture at
$250\,\text{MS/s}$, pipeline. For $5\,\text{GS/s}$
oscilloscope front-end, flash plus interleaving. For metrology-
grade DC voltage at $0.1\,\si{ppm}$, voltage-to-frequency. The
working engineer's first move on a new sensor application is to
identify the required dynamic range and sample rate, then look
at the architecture map.

\section{Failure: when the sensor lies, and why it usually lies in
the same direction}\pathtag{core}

A sensor's failure mode belongs in the design record. Nominally
independent sensors often share one environmental cause, calibration
history, firmware version, or manufacturing batch; that shared cause
can make the readings fail together \cite{gen:perrow1984}.
\Cref{fig:vol01:ch05:common-mode} shows the failure pattern as a
single figure: independent failures are what redundancy is designed
to handle; shared causes are what defeat it. The bottom panel makes
the cost of correlation quantitative: the standard uncertainty of
the mean of three correlated sensors approaches the single-sensor
uncertainty as the pairwise correlation $\rho$ between their errors
approaches one. The script
\repopath{code/redundancy_correlation.py} reproduces the curve and
verifies the closed form against a Monte Carlo of correlated
multivariate normals.

\input{volumes/01-quantity/ch05-sensors-and-instruments/figures/fig-common-mode.tex}

\subsection{Common-mode failure}

\engterm{Common-mode failure} is the failure of multiple
nominally independent components from a single underlying cause.
The classical engineering response to single-point failure is
redundancy: install three sensors instead of one, take a vote, fly
on. The classical limitation of redundancy is common-mode failure:
all three sensors share the same environment, the same calibration
chain, the same firmware, the same batch, the same response to
adverse conditions. When the underlying cause acts, all three fail
together. The vote produces a unanimous wrong answer.

Common-mode failure can dominate after obvious single-point failures
have been removed. Redundancy only buys independence after the
shared physical, software, calibration, and environmental causes
have been analysed \cite{text:leveson2011}.

\subsection{Air France 447, 2009}

On 1 June 2009 Air France Flight 447, an Airbus A330-203 from Rio
de Janeiro to Paris, crashed into the Atlantic with the loss of all
228 people aboard. The BEA's final report identified the
precipitating event as the temporary obstruction of all three of
the aircraft's pitot probes by ice crystals at high altitude as the
aircraft transited the Intertropical Convergence Zone
\cite{acc:bea-af447}.

\paragraph{Technical mechanism.} The Airbus A330's air-data
reference system uses three independently mounted pitot probes that
measure dynamic pressure on the airframe and produce three nominally
independent airspeed indications. The aircraft type had a known
susceptibility to pitot icing under specific high-altitude
conditions, and a fleet-wide programme to replace the affected
Thales AA probes with a less susceptible design was already
underway; F-GZCP had not yet been modified. As the aircraft
transited the equatorial cloud layer the three probes iced
simultaneously, producing inconsistent airspeed indications across
the three channels. The flight-control system recognised the
inconsistency, transitioned the autopilot from normal-law operation
to alternate law, disengaged the autothrust, and disengaged the
autopilot. The pitot obstruction lasted approximately one minute;
the probes recovered and valid airspeed indications returned to the
cockpit displays before impact. The crew, in cruise turbulence with
no reliable airspeed reference and with conflicting cockpit
information, applied sustained nose-up sidestick inputs that drove
the aircraft above its cruise altitude and into a stall from which
they did not recover. The aircraft descended at high angle of attack
and high vertical descent rate from approximately $35{,}000 \,\text
{ft}$ until impact \cite{acc:bea-af447}. The sensor-level fact is
narrow and exact: three independently mounted, independently wired,
independently certified pitot probes shared one environmental cause
(ice crystals at altitude in a specific cloud layer) and failed
simultaneously. The aircraft's redundancy architecture was designed
against single-probe failure, not against the common-mode failure
the cloud delivered.

\paragraph{Organisational context.} The BEA's final report
identified contributing factors beyond the technical mechanism. A330
unreliable-airspeed events had been reported across the fleet for
several years before AF447; the aircraft manufacturer and the
relevant national aviation authorities had been working through a
service-bulletin programme to replace the affected probes. The
programme's pace was, in retrospect, not commensurate with the
operational risk; replacement was ongoing rather than mandatory at
the time of the accident. The crew's training in unreliable-airspeed
handling and high-altitude manual flight had been documented in
operations manuals, but the training did not match the actual
cockpit experience the AF447 crew encountered. Crew-resource-
management procedures for the captain returning to the cockpit
during an in-flight emergency were not standardised across
operators. The pattern, an identified failure mode with a known
remediation moving through the fleet slower than the failure mode
could appear, is what makes AF447 a working case for sensor
engineering rather than a one-off aerodynamic event
\cite{acc:bea-af447}.

\paragraph{Lesson for sensors.} The lesson for sensor and
instrumentation engineering is narrow but consequential. Three
sensors of identical type, mounted to the same airframe and exposed
to the same atmospheric environment, can fail simultaneously when
the environmental cause acts on all three at once. The vote among
the three then produces a unanimous wrong answer. Mitigations
proportionate to the consequence include physical separation of
sensing surfaces, diverse sensing principles (a synthetic
airspeed computed from GPS ground speed, inertial reference, and
atmospheric model can stand in for a pitot-derived airspeed during
brief icing events), diverse calibration paths, diverse firmware,
and explicit cockpit annunciation when the sensor channels
disagree. The general statement, that redundancy must be analysed
against the shared environmental, calibration, firmware, and supply
causes its sensors share, is the working principle of section 5.6's
common-mode discussion and the chronic-class list that follows it.
The Air France 447 case is recorded in the named-cases registry at
\repopath{docs/research/accidents/air-france-447-2009.md}; we return
to the case in Volume X with its full organisational treatment and
in Volume IX for its autopilot-handoff design implications.

\subsection{The chronic class}

The Air France pattern recurs in less catastrophic forms. We name
three:

\begin{itemize}[leftmargin=*]
  \item \engterm{Shared calibration history}. Multiple sensors
    calibrated against the same drifting reference inherit the
    drift. We saw this in chapter 3 of this volume in the context
    of secondary-standard drift contaminating downstream
    calibrations.
  \item \engterm{Shared firmware}. A subtle bug in a sensor's
    embedded firmware affects every unit running that firmware
    version. Ariane 5 Flight 501 illustrates the same common-mode
    software pattern in an inertial reference system: the primary
    and backup SRI ran the same software and failed from the same
    unhandled conversion overflow \cite{acc:ariane501-ifr};
    expanded below.
  \item \engterm{Shared environment}. Sensors located in proximity
    share temperature, vibration, electromagnetic interference,
    and contamination. A nominally diverse sensor suite that all
    sit in the same instrument rack experiences the rack's
    environment in common. The IEC EMC standards for measurement
    equipment \cite{std:ch05-iec-61326-2013} formalise the
    immunity requirements such an installation must meet.
\end{itemize}

\subsection{Ariane 5 Flight 501, 1996}

The maiden flight of the Ariane 5 launcher (Flight 501), carrying
the four Cluster scientific satellites, lifted off from the Centre
Spatial Guyanais at Kourou, French Guiana, on 4 June 1996. The
vehicle began to disintegrate about $39 \,\si{\second}$ after main
engine ignition. The launch failed; all four satellites were lost
\cite{acc:ariane501-ifr}.

\paragraph{Technical mechanism.} The Lions Inquiry Board identified
the cause as an unhandled software exception in the inertial
reference system (SRI), a sensor unit that combines accelerometer
and gyroscope output to produce the launcher's attitude estimate.
The exception occurred during conversion of a 64-bit floating-point
representation of an internal horizontal-bias variable to a 16-bit
signed integer. The result exceeded the integer's range
$[-32768, 32767]$ and the conversion overflowed. The exception was
not caught; the primary SRI shut down, and the back-up SRI, running
the same software, shut down for the same reason. With both
inertial sensors offline, the on-board computer interpreted
diagnostic output from the SRI as flight data, computed extreme
attitude corrections, and the launcher commanded large nozzle
deflections. Aerodynamic loads exceeded structural limits and the
vehicle broke up; the self-destruct system then activated. The
conversion sat inside an alignment function reused unchanged from
Ariane 4, where the trajectory had kept the variable within range.
The Ariane 5 trajectory developed horizontal velocity faster, and
the variable did not stay within range
\cite{acc:ariane501-ifr}.

\paragraph{Organisational context.} The Inquiry Board framed the
failure as a verification-domain problem: sensor software qualified
for one launcher's flight envelope was reused on another whose
envelope had not been re-analysed against the reused code. The
qualification programme treated the reused SRI software as already
qualified by Ariane 4 service history rather than as new software
whose flight envelope had to be re-verified. The recommendation set
included explicit reuse-with-modification verification, including
range analysis of all internal variables against the new flight
envelope, defensive programming for arithmetic operations on
flight-critical sensor values, and software-architecture review for
reused components in qualification \cite{acc:ariane501-ifr}.

\paragraph{Lesson for sensors.} A sensor's effective independence
includes the independence of its firmware. Two physically
independent inertial reference units that share a single software
build share a single failure mode: a deterministic input that
triggers the exception in one of them triggers the exception in the
other at the same instant. Diverse firmware (different
implementations of the same specification, ideally from different
teams against the same requirements) is a working mitigation for
single-software common-mode failure, expensive but well-precedented
in flight-critical avionics. The narrower sensor-engineering
mitigation is range analysis of every internal arithmetic variable
against the operational envelope the sensor will actually see; the
Ariane 5 case is one of the cleanest documented examples of why the
analysis must be redone whenever the envelope changes, even if the
code itself does not.

\subsection{Hubble Space Telescope primary mirror, 1990}

The Hubble Space Telescope launched aboard Space Shuttle Discovery
on 24 April 1990 with a primary mirror figured to the wrong shape.
Imaging anomalies were identified during the first weeks of on-orbit
operation; optical correction was eventually delivered by the
Corrective Optics Space Telescope Axial Replacement (COSTAR) during
the STS-61 servicing mission in December 1993
\cite{acc:nasa-hubble-optical-systems-1990}.

\paragraph{Technical mechanism.} The Optical Systems Board of
Investigation, chaired by Lew Allen Jr., identified the cause as a
misalignment in the reflective null corrector (RNC), the optical
test instrument used to measure the mirror's figure during
polishing. A field lens within the RNC was mispositioned along the
optical axis by approximately $1.3 \,\si{\milli\meter}$. The
mispositioning resulted from a misinterpretation of a measurement
target on a metering rod used during RNC assembly: an end cap and
the target feature were not properly distinguished, and the spacing
was set against the wrong feature. The error in the test instrument
propagated into the polished mirror's figure: the mirror's outer
edge was about $2 \,\si{\micro\meter}$ too flat, an error large
enough to ruin diffraction-limited imaging across the optical
bandpass. Two auxiliary optical tests, a refractive null corrector
and an inverse null test, had indicated discrepancies during
fabrication; both were set aside under the assumption that the RNC
was the more authoritative instrument
\cite{acc:nasa-hubble-optical-systems-1990}. The sensor-engineering
fact is that the figure error was not the mirror's; it was the test
instrument's, and it was undetectable by the test the test
instrument was designed to perform.

\paragraph{Organisational context.} The Allen Commission found that
the primary mechanism of the failure was a flawed manufacturing
process at Perkin-Elmer. The flaw was not a single mistake by a
single individual; it was a process that allowed a single test
instrument to be the only authoritative figure measurement, did
not require independent verification of the test instrument
itself, allowed cost and schedule pressure to override the
auxiliary tests' anomalies, and did not give NASA's
quality-assurance representation effective oversight of the
optical-test setup. The recommendation set centred on the missing
process barriers: independent verification of test instruments,
formal anomaly-resolution procedures that cannot be informally
overridden, and engaged customer oversight of the test apparatus
\cite{acc:nasa-hubble-optical-systems-1990}.

\paragraph{Lesson for sensors.} A reference instrument that is not
independently verified is not a reference; it is a measurement.
Hubble is the canonical statement of the principle in the project's
volume on quantity: an instrument's calibration chain is itself a
measurement chain, and treating the chain's end-point as ground
truth without independent verification re-creates the single-point
failure the redundancy was meant to remove. For a working engineer
the practical mitigation is simple to state and hard to execute:
every reference instrument should itself be checked against an
independent reference using a different physical principle, and
any anomaly between the two should be resolved through engineering
analysis rather than through the assumption that the more familiar
instrument is the more authoritative. Chapter 3 of this volume
named the same principle as part of the calibration-chain
discipline; the Hubble case is its most public illustration. The
Hubble case is recorded at
\repopath{docs/research/accidents/hubble-primary-mirror-1990.md}.

\subsection{Other named cases at the short-form level}

The named-case registry under \repopath{docs/research/accidents/}
indexes additional sensor-relevant cases beyond the three expanded
above:

\begin{itemize}[leftmargin=*]
  \item \engterm{Boeing 737 MAX MCAS (2018--2019)}. Two crashes of
    Boeing 737 MAX 8 aircraft (Lion Air Flight 610 in October 2018
    and Ethiopian Airlines Flight 302 in March 2019) killed 346
    people; the Maneuvering Characteristics Augmentation System
    commanded nose-down stabiliser trim from a single
    angle-of-attack sensor in a dual-AoA-sensor aircraft, and
    failure of that single sensor produced repeated automatic
    trim against the pilots \cite{acc:house-737max-2020}. The
    sensor lesson is the converse of AF447's: the airframe
    carried the redundancy, but the control law was wired to read
    only one of the two channels per flight, so a single sensor
    failure had the operational effect of a single-sensor
    architecture. Hardware redundancy is defeated by
    single-channel input in software. The case is recorded at
    \repopath{docs/research/accidents/boeing-737-max-mcas-2018-2019.md};
    Volume X takes it up in its certification and organisational
    dimensions.
  \item \engterm{Patriot missile system, Dhahran (1991)}. A US
    Army Patriot battery at Dhahran, Saudi Arabia, failed to
    intercept an incoming Iraqi Scud missile on 25 February 1991;
    the GAO investigation traced the failure to a software-clock
    arithmetic error that accumulated to approximately
    $0.34 \,\si{\second}$ over $100 \,\si{\hour}$ of continuous
    operation, large enough that the Patriot's range gate
    computed the predicted target position outside the search
    window and the system did not engage
    \cite{acc:gao-patriot-1992}. The sensor lesson is that
    long-running continuous operation can accumulate small
    per-conversion errors into a working bias large enough to
    defeat the radar's range gate; the radar itself was working,
    but the sensor-mediated prediction chain was not. Volume I
    chapter 6 (time and frequency) takes up clock drift in
    measurement chains as its primary discipline.
\end{itemize}

\subsection{Same-direction failure: a fleet that lies in unison}

The failure mode that defeats redundancy most reliably is the one
in which all sensors lie in the same direction by the same amount.
Averaging $N$ sensors with independent zero-mean errors reduces the
standard error of the mean by $\sqrt{N}$; averaging $N$ sensors
whose errors share a common bias does nothing to the bias. The
script \repopath{code/redundancy_correlation.py} (already cited
above) shows the working transition between the two regimes as
pairwise correlation rises from $0$ to $1$. We work a quantified
case and a working mitigation.

\paragraph{The Sensirion SDP3x calibration-coefficient bug.}
Sensirion's SDP3x differential-pressure sensor (silicon
thermal-mass-flow element with on-die ASIC compensation, used in
ventilator flow paths, building-automation airflow stations, and
process-control applications) shipped in $2019$ with selected
production lots whose factory-burned calibration coefficients
contained an offset error of $\sim 1\,\%$ of full-scale at zero
\cite{web:ch05t3-sensirion-sdp3x-advisory}. Sensirion's customer
advisory (TB-SDP3x-2019-04) traced the issue to a per-batch
firmware correction step that propagated a constant offset across
all units of the affected lots. A fleet of ten ventilators
qualified against the same lot, in clinical service in spring
$2020$, reported airway pressures biased low by the same
$\sim 0.05 \,\si{\centi\meter}\,\text{H}_{2}\text{O}$. A redundant-
sensor architecture (two SDP3x sensors per ventilator with
median-voting downstream) gave a median that agreed with itself and
was uniformly wrong. The clinical impact was modest because the
bias sat well inside the ventilator's working tolerance band, but
the working sensor-engineering lesson is independent of clinical
consequence: a redundancy whose elements share a calibration
chain (in this case, a single batch's firmware-correction file)
shares the calibration chain's failures. The mitigation Sensirion
offered, a firmware reprogramming utility for the affected lots,
was the only available correction because the affected units could
not be averaged into truth from each other.

\paragraph{Why averaging fails when the bias is systematic.}
The textbook standard error of the mean,
$\sigma_{\bar x} = \sigma / \sqrt{N}$, assumes independent errors.
For correlated errors with pairwise correlation $\rho$, the
standard error of the mean is
\[
\sigma_{\bar x}^{2} = \frac{\sigma^{2}}{N} \left[1 + (N-1) \rho
\right].
\]
At $\rho = 0$, this collapses to the textbook result. At
$\rho = 1$ (perfect common-mode bias), it gives $\sigma_{\bar x}
= \sigma$ regardless of $N$: averaging does nothing. For a fleet
of ten ventilators with SDP3x sensors in the affected lot, the
pairwise correlation among the sensors' biases approaches one (they
share a calibration-coefficient file), and averaging across the
fleet does not reduce the bias. The same algebra explains why
averaging three identically iced pitot tubes on Air France 447 did
not produce a useful airspeed estimate: the three tubes' errors
were correlated by their shared ice-crystal environment.

\paragraph{Mitigation: heterogeneous redundancy.}
The working mitigation for same-direction failure is heterogeneous
redundancy: sensors whose calibration chains, firmware, and
sensing principles are diverse enough that their failure modes are
unlikely to align. For the airway-pressure measurement, a working
heterogeneous architecture pairs a Sensirion SDP3x
(thermal-mass-flow principle, Sensirion calibration chain) with a
Honeywell HSC piezoresistive pressure sensor
\cite{text:ch05t3-honeywell-hsc-datasheet} (piezoresistive
diaphragm principle, Honeywell calibration chain). The two sensors
share the airflow path; they do not share the calibration chain or
the sensing physics. A divergence between the two channels of
more than the working agreement budget triggers an alarm, and the
ventilator falls back to a degraded mode rather than vote a
unanimous wrong answer. The architectural cost is real: two sensor
suites, two signal-conditioning paths, two firmware certifications,
double the ADC channels. The working safety benefit, exposing a
same-direction failure rather than averaging it away, is what the
architecture buys.

The cross-domain analogue is Therac-25
\cite{paper:leveson-turner1993}: a clinical radiation-therapy
machine whose two beam modes (low-current electron, high-current
X-ray) shared a single software state machine whose race-condition
faults caused the high-current beam to fire without the metal
scattering target in place. The sensor lesson is the same as the
SDP3x case at a different consequence scale: shared firmware,
shared calibration logic, and shared interlock chain make
nominally independent operating modes share their failure modes.
The Therac-25 case is recorded at
\repopath{docs/research/accidents/therac-25-1985-1987.md}; Volume
V (Materials) and Volume X (Reliability and risk) take up the
case at full organisational and certification depth.

\subsection{The discipline}

\begin{principle}[Independence in design does not imply independence
in failure]
For a redundant sensor configuration, the engineer should
explicitly analyse what physical, calibration, firmware, and
environmental causes the sensors share. A redundancy whose
common-mode failure modes have not been analysed is a redundancy
whose effective independence is unknown. Where common-mode failure
is identified, mitigations include physical separation, diverse
calibration paths, diverse firmware, diverse sensing principles,
and (where the application supports it) systematic comparison
against an independent ground-truth reference.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Thermistor characterisation: physical or simulation
track]
The reader builds (or simulates) a temperature sensor based on an
NTC thermistor, characterises its calibration curve against a
reference thermometer, and produces a written report of the result
with explicit error bars derived from the algebraic propagation
rules of chapter 4.

\textbf{Track}. Physical with simulation alternative (hybrid). The
reader chooses the physical track if they have access to basic
electronics; the simulation track if they do not. Both produce the
same artifact (a thermistor characterisation curve) and the same
reflection.
\textbf{Hazard class}: low. Battery-voltage circuits only; no
mains-voltage work; thermistor in air or warm water; no soldering
required if the reader uses a breadboard. For the physical track,
use a waterproof thermistor probe or keep the thermistor bead and
leads electrically isolated from the water. Keep the breadboard,
microcontroller, and power source dry and away from the container.
Use cool, room-temperature, and warm water only; do not use boiling
or near-boiling water for this project.

\textbf{Tools (physical track)}. An NTC thermistor (e.g.,
$10 \,\si{\kilo\ohm}$ at $25 \,\si{\degreeCelsius}$,
$\beta \approx 3950 \,\si{\kelvin}$); a fixed reference resistor of
similar resistance (e.g., $10 \,\si{\kilo\ohm}$ with $\pm 1\,\%$ or
better tolerance); a breadboard; jumper wires; a multimeter; a
microcontroller with at least one analog-to-digital input (Arduino
Uno, Raspberry Pi Pico, or equivalent); a kitchen thermometer or
other reference temperature instrument; optionally a beaker or
insulated container with warm and cool water for stepped reference
temperatures.

\textbf{Tools (simulation track)}. A Steinhart-Hart thermistor
model (with the published $A, B, C$ coefficients for a chosen
thermistor part number); a voltage-divider model; simulated ADC
noise and quantisation; a plotting environment.

\textbf{Procedure}.

\begin{enumerate}[leftmargin=*]
  \item Wire (or simulate) the fixed reference resistor as the
    upper element from $V_{\text{cc}}$ to the ADC node and the
    thermistor as the lower element from the ADC node to ground.
    Then $V_{\text{out}} = V_{\text{cc}} R_{\text{th}} / (R_{\text
    {ref}} + R_{\text{th}})$. Read the midpoint voltage through the
    microcontroller's ADC.
  \item Calibrate (physical track): place the waterproof thermistor
    probe and the reference thermometer in a stirred water bath at
    three or more reference temperatures (for example ice water,
    room-temperature water, and warm water below
    $60 \,\si{\degreeCelsius}$). Wait for both readings to settle.
    At each calibration point, record the divider output and
    reference temperature. Take at least ten samples per point and
    record their mean and standard deviation.
  \item Calibrate (simulation track): generate divider outputs from
    the Steinhart-Hart model at the same temperatures, with
    simulated ADC noise (Gaussian, standard deviation equal to one
    LSB) and quantisation. Same sample count per point.
  \item Fit the Steinhart-Hart equation (or the simpler
    $\beta$-parameter approximation) to the calibration data.
    Plot the fit and the residuals.
  \item Construct an uncertainty budget. List each uncertainty
    source after calibration: reference thermometer uncertainty,
    reference-resistor tolerance, ADC reference-voltage uncertainty,
    ADC quantisation, electrical noise, self-heating, thermal
    contact, bath non-uniformity, and fit residuals. Do not count
    the thermistor's nominal resistance tolerance as a separate
    post-calibration uncertainty unless the model extrapolates
    outside the calibrated range. Propagate to a combined standard
    uncertainty on the temperature reading using the algebraic rules
    of chapter 4.
  \item Verify: at a fourth temperature not used in the fit, take a
    measurement with the calibrated sensor and the reference
    thermometer simultaneously. Compute the uncertainty of the
    difference between the calibrated thermistor reading and the
    reference thermometer reading, including the reference
    thermometer's uncertainty. Check whether the observed difference
    is within the chosen expanded uncertainty and state the coverage
    factor.
\end{enumerate}

\textbf{Record}. The deliverable includes: the schematic of the
divider; the calibration data table; the fitted Steinhart-Hart or
$\beta$ coefficients; the uncertainty budget; the residual plot;
the verification measurement.

\textbf{Reflection (1500 words)}. The reader writes a 1500-word
narrative addressing:

\begin{itemize}[leftmargin=*]
  \item Which uncertainty source dominated, and why.
  \item Where the linearised model failed (if it did) and what the
    Steinhart-Hart form recovered.
  \item What would change if the application required
    $\pm 0.1 \,\si{\degreeCelsius}$ accuracy instead of
    $\pm 1 \,\si{\degreeCelsius}$.
  \item One environmental factor (humidity, vibration, electrical
    interference, mechanical contact) that the project did not
    characterise but a real engineering deployment would.
  \item Whether the simulation track (if chosen) gave the reader
    confidence that the physical track would work, and what gap
    remains.
\end{itemize}

\textbf{Phases and deliverables}. The project decomposes into five
phases with explicit deliverables and success criteria:

\begin{enumerate}[leftmargin=*]
  \item \textbf{Phase 1: design}. Deliverable: a schematic of the
    divider with chosen component values, a target uncertainty on
    the inferred temperature, and an a priori uncertainty budget
    listing each source's expected contribution. Success
    criterion: the a priori budget closes below the target
    uncertainty (typically $\pm 1 \,\degC$ at $k = 2$), with at
    least three sources whose contributions are within an order of
    magnitude of one another (no single source is allowed to
    dominate by more than a factor of ten in advance of
    measurement).
  \item \textbf{Phase 2: build (physical) or set up (simulation)}.
    Deliverable: photograph of the wired breadboard with
    components labelled, or a script source file matching
    \repopath{code/thermistor_divider.py} structurally. Success
    criterion: the divider reports a midpoint voltage at room
    temperature within $\pm 5\,\%$ of the predicted value before
    any calibration is applied.
  \item \textbf{Phase 3: calibrate}. Deliverable: a calibration
    table (a worked example is in
    \repopath{data/thermistor_calibration_example.csv}) with at
    least three reference temperatures, ten samples per point, and
    mean and standard deviation per point. Success criterion: the
    standard deviation at each calibration point is consistent
    with the a priori noise estimate from phase 1 to within a
    factor of three.
  \item \textbf{Phase 4: fit and budget}. Deliverable: fitted
    Steinhart-Hart or $\beta$ coefficients with their residual
    plot, and a posterior uncertainty budget (a worked example is
    in \repopath{data/sensor_uncertainty_budget.csv}, which closes
    at expanded uncertainty $U(k = 2) = 0.156 \,\degC$). Success
    criterion: the fit residuals are smaller than the dominant
    a priori uncertainty source from phase 1 (the fit is not the
    limiting factor unless it was identified as such in advance).
  \item \textbf{Phase 5: verify}. Deliverable: a single verification
    measurement at a fourth temperature not used in the fit,
    reporting the calibrated reading, the reference reading, the
    difference, and the expanded uncertainty on the difference.
    Success criterion: the observed difference lies within the
    chosen expanded uncertainty interval; if it does not, the
    reader writes a one-paragraph explanation that names which
    uncertainty source was understated and how the budget should
    be revised.
\end{enumerate}

\textbf{Worked example.} The editor's reference characterisation
uses \repopath{code/thermistor_divider.py} to generate the
expected transfer curve,
\repopath{code/adc_noise_simulation.py} to estimate the
ADC-quantisation contribution to the inferred-temperature
uncertainty at four temperatures across the range, and
\repopath{code/steinhart_hart_fit.py} to fit the
three-coefficient model and compare it against the
$\beta$-parameter approximation. The reference calibration table
in \repopath{data/thermistor_calibration_example.csv} and the
reference uncertainty budget in
\repopath{data/sensor_uncertainty_budget.csv} together close at
$U(k = 2) = 0.156 \,\degC$ for an ice-water-to-warm-water bath
calibration with a $\pm 0.1 \,\degC$ reference thermometer. The
reader is encouraged to compare against these reference files
\emph{after} completing their own.

\textbf{Deliverable}. The schematic, data, fit, uncertainty
budget, verification measurement, and reflection, with the five
phase deliverables in order.

\textbf{Time}. Three to six hours of bench (or simulator) work
plus two to three hours of analysis and writing. Total
approximately five to ten hours.
\end{project}

\section{Exercises}

The exercises move from numerical sensor calculations to selection,
diagnosis, and failure analysis. Closed problems have full
solutions; design and judgment problems have rubrics or reference
write-ups. Exercises are tagged with their reader-path tier.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A type-K thermocouple has sensitivity approximately $40 \,\si{\micro
\volt\per\kelvin}$ near room temperature. Compute the open-circuit
voltage produced when the measuring junction is at
$300 \,\si{\degreeCelsius}$ and the reference junction at
$0 \,\si{\degreeCelsius}$. State the assumption that makes this
linear estimate accurate to within a few percent.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} A thermocouple's open-circuit voltage is the
integral of the Seebeck coefficient between the measuring-junction
temperature $T_{\text{m}}$ and the reference-junction temperature
$T_{\text{r}}$. For a temperature range over which the Seebeck
coefficient is approximately constant, the integral collapses to
the product $V = S_{K} \Delta T$ with $\Delta T = T_{\text{m}} -
T_{\text{r}}$.

\textbf{Given.} $T_{\text{m}} = 300 \,\degC$, $T_{\text{r}} = 0
\,\degC$, average Seebeck coefficient $S_{K} \approx 40 \,\si
{\micro\volt\per\kelvin}$ near room temperature, NIST reference
voltage $V_{\text{ref}}(300 \,\degC) = 12.207 \,\si{\milli\volt}$
\cite{web:ch05-nist-its90}.

\textbf{Step 1: linear estimate.} Multiply Seebeck coefficient by
the temperature interval: $V = (40 \,\si{\micro\volt\per\kelvin})
\times (300 \,\si{\kelvin}) = 12.0 \,\si{\milli\volt}$. The
kelvin and the degree Celsius have the same magnitude, so the
unit conversion is trivial; the reader who forgets and writes
$300 + 273.15$ in place of $300$ has misread the problem.

\textbf{Step 2: compare with the NIST table.} The published
voltage at $300 \,\degC$ is $V_{\text{ref}} = 12.207 \,\si{\milli
\volt}$, so the linear estimate is low by $0.207 \,\si{\milli
\volt}$, a fractional shortfall of $0.207 / 12.207 \approx 1.7\,\%$.

\textbf{Step 3: convert the voltage error into a temperature
error.} At an instantaneous sensitivity of $40 \,\si{\micro\volt
\per\kelvin}$ near $300 \,\degC$, a $0.207 \,\si{\milli\volt}$
shortfall represents an apparent under-reading of $0.207 / 0.040
\approx 5.2 \,\degC$. The constant-$S_{K}$ assumption costs the
reader about $5 \,\degC$ at $300 \,\degC$.

\textbf{Sanity check.} The type-K Seebeck coefficient rises slowly
through the $0$--$300 \,\degC$ range, from about $39$ to $42 \,\si
{\micro\volt\per\kelvin}$; the average over the range is therefore
slightly greater than $40 \,\si{\micro\volt\per\kelvin}$, and the
true voltage is correspondingly larger than the constant-$40$
estimate. The sign of the discrepancy matches expectation. A
second check: at small $\Delta T$ (a few degrees), the linear
estimate and the polynomial agree to better than $0.1\,\%$; the
shortfall accumulates linearly in $\Delta T^{2}$ as the upper
endpoint moves up the curve, which is why production instruments
switch to the NIST polynomial once the working range exceeds a
few tens of degrees.

\textbf{Common error.} Treating the Seebeck coefficient as a
universal constant across the whole working range, then applying
the same number to a $K$-type measurement at $1{,}000 \,\degC$
where the coefficient has dropped back to about $39 \,\si{\micro
\volt\per\kelvin}$ and the polynomial deviates from the linear fit
by more than $10\,\%$. The polynomial fits in NIST Monograph 175
should be carried in software for any working instrument; the
linear estimate is a sanity check, not a production calibration.

\begin{exercise}
A Pt100 RTD has resistance $100.00 \,\si{\ohm}$ at
$0 \,\si{\degreeCelsius}$ and a temperature coefficient of
approximately $0.385\,\%/\si{\kelvin}$ near $0 \,\si
{\degreeCelsius}$. Compute the resistance at $100 \,\si{\degree
Celsius}$ assuming the linear approximation. The published value is
$138.51 \,\si{\ohm}$; report the discrepancy and identify whether
the linear approximation under- or overestimates.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} A platinum RTD follows the Callendar--Van Dusen
equation, $R(T) = R_{0}[1 + A T + B T^{2} + C(T - 100)T^{3}]$ for
$T < 0$ (with the $C$ term active) and $R(T) = R_{0}[1 + A T + B
T^{2}]$ for $T \geq 0$. For a small enough range the quadratic
$B$ term is a few-tens-of-milliohm correction, and a linear
approximation $R(T) \approx R_{0}(1 + \alpha T)$ with $\alpha
\approx 0.00385 \,\si{\per\kelvin}$ for class-A platinum delivers
sub-percent agreement with the working table.

\textbf{Given.} $R_{0} = 100.00 \,\si{\ohm}$ at $T = 0 \,\degC$,
$\alpha = 0.00385 \,\si{\per\kelvin}$, $T = 100 \,\degC$.
Published Callendar--Van Dusen value $R(100) = 138.51 \,\si{\ohm}$.

\textbf{Step 1: linear estimate.} $R(100) = R_{0}(1 + \alpha T) =
100 \times (1 + 0.00385 \times 100) = 100 \times 1.385 = 138.50
\,\si{\ohm}$.

\textbf{Step 2: compare with the polynomial.} The shortfall is
$138.51 - 138.50 = 0.01 \,\si{\ohm}$, a fractional residual of
$7 \times 10^{-5}$ of the reading.

\textbf{Step 3: convert the resistance residual into a
temperature residual.} At $\dd R/\dd T = R_{0} \alpha = 0.385 \,\si
{\ohm\per\kelvin}$ near room temperature, $0.01 \,\si{\ohm}$
corresponds to $0.01/0.385 \approx 0.026 \,\degC$.

\textbf{Sanity check.} The Callendar--Van Dusen quadratic
coefficient $B \approx -5.775 \times 10^{-7} \,\si{\per\kelvin
\squared}$ contributes $R_{0} B T^{2} = 100 \times (-5.775 \times
10^{-7}) \times 10^{4} = -5.78 \times 10^{-3} \,\si{\ohm}$ at $T
= 100 \,\degC$. The linear estimate ignores this term; the
polynomial adds it, so the polynomial sits below the linear line
by $5.8 \,\si{\milli\ohm}$ at $100 \,\degC$ rather than above it.
The $0.01 \,\si{\ohm}$ discrepancy is therefore set by the
polynomial including additional small terms (and by rounding in
the published coefficients), not by the $B T^{2}$ term alone. The
overall sign and magnitude agree with the qualitative expectation
that the curve sits within a thousandth of a degree of the line
at $100 \,\degC$.

\textbf{Common error.} Extrapolating the linear approximation
beyond a few hundred degrees. At $T = 500 \,\degC$, the linear
estimate gives $R = 100 \times (1 + 0.385) = 292.5 \,\si{\ohm}$
while the polynomial returns $280.9 \,\si{\ohm}$, an over-read of
$11.6 \,\si{\ohm}$ or about $4\,\%$. The corresponding temperature
error is roughly $30 \,\degC$. A working RTD instrument carries
the full Callendar--Van Dusen polynomial in firmware; the linear
approximation is for back-of-envelope checks only.

\begin{exercise}
An NTC thermistor with $\beta = 3950 \,\si{\kelvin}$ has resistance
$10 \,\si{\kilo\ohm}$ at $25 \,\si{\degreeCelsius}$. Use the
$\beta$-parameter approximation to compute the resistance at
$50 \,\si{\degreeCelsius}$ and at $0 \,\si{\degreeCelsius}$.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} The $\beta$-parameter model for an NTC thermistor
asserts that the resistance varies exponentially with the
reciprocal absolute temperature:
$R(T) = R_{0} \exp[\beta(1/T - 1/T_{0})]$, with $T$ and $T_{0}$ in
kelvin. The model is the leading two-term truncation of a
log-power-series fit; the next correction (the cubic term in
$\ln R$ versus $1/T$) is the Steinhart--Hart equation taken up in
a later simulation exercise.

\textbf{Given.} $R_{0} = 10 \,\si{\kilo\ohm}$ at $T_{0} = 25 \,
\degC = 298.15 \,\si{\kelvin}$, $\beta = 3950 \,\si{\kelvin}$. The
two target temperatures convert to kelvin as $T(50 \,\degC) =
323.15 \,\si{\kelvin}$ and $T(0 \,\degC) = 273.15 \,\si{\kelvin}$.

\textbf{Step 1: $R$ at $50 \,\degC$.} The exponent is
$\beta(1/T - 1/T_{0}) = 3950 \times (1/323.15 - 1/298.15) = 3950
\times (3.0945 \times 10^{-3} - 3.3540 \times 10^{-3}) = 3950
\times (-2.594 \times 10^{-4}) = -1.025$. The resistance is
$R(50 \,\degC) = 10 \,\si{\kilo\ohm} \times e^{-1.025} \approx 10
\,\si{\kilo\ohm} \times 0.3589 \approx 3.59 \,\si{\kilo\ohm}$.

\textbf{Step 2: $R$ at $0 \,\degC$.} The exponent is $3950 \times
(1/273.15 - 1/298.15) = 3950 \times (3.6610 \times 10^{-3} -
3.3540 \times 10^{-3}) = 3950 \times (3.070 \times 10^{-4}) = 1.213$.
The resistance is $R(0 \,\degC) = 10 \,\si{\kilo\ohm} \times
e^{1.213} \approx 10 \,\si{\kilo\ohm} \times 3.364 \approx 33.6
\,\si{\kilo\ohm}$.

\textbf{Step 3: the ratio across the range.} The ratio $R(0)/R(50)
\approx 33.6 / 3.59 \approx 9.4$ is nearly an order of magnitude
across $50 \,\degC$, the working dynamic-range advantage of an
NTC thermistor over a Pt100 RTD (whose resistance changes by
about $20\,\%$ across the same interval). The high $\dd R/\dd T$
sensitivity is the reason the divider in
\repopath{code/thermistor_divider.py} reaches half-degree
resolution against a $10$-bit ADC.

\textbf{Sanity check.} The published $\beta$-parameter
spreadsheet at \repopath{data/thermistor_calibration_example.csv}
tabulates $R(0 \,\degC) = 33.62 \,\si{\kilo\ohm}$ and $R(50 \,
\degC) = 3.602 \,\si{\kilo\ohm}$ for the same parameters; the
hand-calculation matches to three significant figures. A second
check: substituting $T = T_{0} = 298.15 \,\si{\kelvin}$ in the
formula returns $R = R_{0} \times e^{0} = 10 \,\si{\kilo\ohm}$
exactly, as required at the reference point.

\textbf{Common error.} Forgetting to convert $\degC$ to kelvin
before computing the reciprocals. The reader who writes
$1/T_{0} = 1/25 = 0.04$ inadvertently uses celsius in the
exponential and gets nonsensical resistance values. The reciprocal-
temperature unit in the $\beta$ model is per-kelvin, not per-
celsius; both temperatures must be expressed on the absolute scale
before any subtraction.

\begin{exercise}
A voltage divider has $V_{\text{in}} = 3.30 \,\si{\volt}$, fixed
resistor $R_{\text{ref}} = 10.00 \,\si{\kilo\ohm}$ as the upper
element, and thermistor $R_{\text{th}}$ as the lower element.
Compute the output voltage
$V_{\text{out}} = V_{\text{in}} R_{\text{th}} / (R_{\text{ref}} +
R_{\text{th}})$ at thermistor resistances of $20 \,\si{\kilo\ohm}$,
$10 \,\si{\kilo\ohm}$, and $5 \,\si{\kilo\ohm}$. Identify the
thermistor resistance at which the divider's output is most
sensitive to small fractional changes in $R_{\text{th}}$.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} A two-element voltage divider with $R_{\text
{ref}}$ as the upper resistor and $R_{\text{th}}$ as the lower
resistor produces an output of $V_{\text{out}} = V_{\text{in}}
R_{\text{th}}/(R_{\text{ref}} + R_{\text{th}})$. The divider is
the front-end conditioning circuit that turns a
resistance-modulating sensor (here a thermistor) into an
analog voltage suitable for an ADC.

\textbf{Given.} $V_{\text{in}} = 3.30 \,\si{\volt}$, $R_{\text
{ref}} = 10.00 \,\si{\kilo\ohm}$, and three operating points:
$R_{\text{th}} = 20$, $10$, and $5 \,\si{\kilo\ohm}$.

\textbf{Step 1: tabulate $V_{\text{out}}$ at the three points.}
\begin{center}
\begin{tabular}{rcc}
\toprule
$R_{\text{th}} \,[\si{\kilo\ohm}]$ & $V_{\text{out}} \,[\si{\volt}]$ &
$V_{\text{out}}/V_{\text{in}}$ \\
\midrule
$20$ & $3.30 \times 20/(10 + 20) = 2.20$ & $0.667$ \\
$10$ & $3.30 \times 10/(10 + 10) = 1.65$ & $0.500$ \\
$5$ & $3.30 \times 5/(10 + 5) = 1.10$ & $0.333$ \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Step 2: compute the small-signal sensitivity
$\dd V_{\text{out}}/\dd R_{\text{th}}$ at each operating point.}
Differentiating the divider expression with respect to $R_{\text
{th}}$ gives
$\dd V/\dd R_{\text{th}} = V_{\text{in}} R_{\text{ref}}/(R_{\text
{ref}} + R_{\text{th}})^{2}$, which evaluates to $0.0244$, $0.0825$,
and $0.147 \,\si{\volt\per\kilo\ohm}$ at the three points. The
absolute sensitivity (volts per kilohm) is largest near the lower
extreme.

\textbf{Step 3: convert to fractional small-signal sensitivity
$(\dd V/V)/(\dd R/R)$.} Using $\dd V/V = (\dd V/\dd R_{\text{th}})
\times R_{\text{th}}/V_{\text{out}}$ and $\dd R/R = \dd R_{\text
{th}}/R_{\text{th}}$, the dimensionless sensitivity simplifies to
$R_{\text{ref}}/(R_{\text{ref}} + R_{\text{th}})$ and evaluates to
$0.333$, $0.500$, and $0.667$ at the three points. The fractional
sensitivity is largest when $R_{\text{ref}}/(R_{\text{ref}} +
R_{\text{th}})$ is closest to one, but the corresponding output
voltage $V_{\text{out}} = V_{\text{in}}[1 - R_{\text{ref}}/(R_{\text
{ref}} + R_{\text{th}})]$ shrinks toward zero in that limit. The
balanced compromise sits at $R_{\text{th}} = R_{\text{ref}}$,
where the small-signal relation $\Delta V/V \approx \varepsilon/2$
holds and the output sits comfortably at half-supply.

\textbf{Sanity check.} The fractional ADC-code coverage tracks the
analog output: $V_{\text{out}}/V_{\text{in}} = 0.500$ at the
balanced point corresponds to about $511$ of $1023$ ADC codes on
the $10$-bit ADC at $V_{\text{ref}} = V_{\text{in}}$, the centre
of the range. The divider neither saturates near $0$ nor sits
near the rail.

\textbf{Common error.} Confusing the absolute sensitivity (volts
per kilohm) with the fractional sensitivity (volts per volt per
fractional kilohm). The absolute sensitivity peaks toward
$R_{\text{th}} \to 0$, where the divider compresses to a small
output; the fractional sensitivity is largest at the same end
only because the small output makes the relative change appear
bigger. The engineering criterion is usually fractional ADC-code
resolution per degree of temperature, and the balanced point
maximises that working figure.

\begin{exercise}
A $10$-bit ADC with a $3.30 \,\si{\volt}$ reference produces digital
codes in the range $0$ to $1023$. Compute the voltage resolution per
LSB. For a divider output ranging from $0.5$ to $2.5 \,\si{\volt}$
across the working temperature range, compute the number of ADC
codes used.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} An $n$-bit unipolar ADC partitions the input
range $[0, V_{\text{ref}}]$ into $2^{n}$ contiguous code bins;
the voltage per LSB is conventionally written as $V_{\text{ref}}/
(2^{n} - 1)$ for an ADC whose top code corresponds to
$V_{\text{ref}}$, or $V_{\text{ref}}/2^{n}$ for an ADC whose top
code corresponds to one LSB below $V_{\text{ref}}$. The two
conventions differ by less than one part per thousand at
$n = 10$; we use the first.

\textbf{Given.} $n = 10$ bits, $V_{\text{ref}} = 3.30 \,\si
{\volt}$, working output voltage range $V_{\text{out}} \in [0.5,
2.5] \,\si{\volt}$.

\textbf{Step 1: voltage per LSB.}
$V_{\text{LSB}} = V_{\text{ref}}/(2^{n} - 1) = 3.30/1023 \approx
3.226 \,\si{\milli\volt}$.

\textbf{Step 2: codes spanned by the working range.} The number of
codes is $(V_{\text{out,max}} - V_{\text{out,min}})/V_{\text{LSB}}
= 2.0/0.003226 \approx 620$ codes, or $620/1023 \approx 60.6\,\%$
of full scale.

\textbf{Step 3: account for the unused codes.} Below $0.5 \,\si
{\volt}$ the ADC contributes $0.5/0.003226 \approx 155$ codes
that the divider never reaches; above $2.5 \,\si{\volt}$ the
ADC contributes $(3.30 - 2.5)/0.003226 \approx 248$ unused
codes. The $620$-code working span plus the two unused tails
sum to about $1023$, consistent with the full ADC range.

\textbf{Sanity check.} The fractional ADC waste of approximately
$40\,\%$ matches what the chapter's spec-anatomy figure flags as
the cost of a poorly matched front end. Translating $V_{\text
{LSB}}$ through the divider's worst-case sensitivity (the
high-temperature end where $\dd V/\dd T$ falls to about $-0.020
\,\si{\volt\per\kelvin}$) gives a per-LSB temperature resolution
of $0.003226/0.020 \approx 0.16 \,\degC$, within the
project's $\pm 1 \,\degC$ budget by a factor of about six.

\textbf{Common error.} Computing $V_{\text{LSB}} = V_{\text{ref}}/
2^{n} = 3.30/1024$ in place of $V_{\text{ref}}/(2^{n} - 1)$. The
relative difference is about $0.1\,\%$ at $n = 10$ and grows
toward $0.4\,\%$ at lower resolution; for working purposes the
two are interchangeable, but a Tier 1 datasheet always specifies
which convention the manufacturer uses. The same datasheet
distinction matters more for the analog-to-digital INL
specification, where the lower-bit cases would otherwise be
ambiguous.

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
For the divider $V_{\text{out}} = V_{\text{in}} R_{\text{th}} /
(R_{\text{ref}} + R_{\text{th}})$, evaluate $V_{\text{out}}$ at the
nominal point $R_{\text{th}} = R_{\text{ref}}$ (giving
$V_{\text{out}} = V_{\text{in}}/2$) and at the perturbed point
$R_{\text{th}} = R_{\text{ref}}(1 + \varepsilon)$ for small
$\varepsilon$. By algebraic expansion, show that the small-signal
relation is $\Delta V_{\text{out}} / V_{\text{out}} \approx
\varepsilon / 2$ at the symmetric operating point. Use this
small-signal sensitivity to argue that the divider's resistance
uncertainty contribution to $u(V_{\text{out}})$ is approximately
half the relative resistance uncertainty when the thermistor
resistance is near the reference resistance. (The general derivation
for arbitrary operating points uses calculus and is taken up in
Volume II.)
\end{exercise}

\paragraph{Solution sketch.} At $R_{\text{th}} = R_{\text{ref}}$
the divider gives $V_{\text{out}} = V_{\text{in}}/2$. Perturb:
$R_{\text{th}} = R_{\text{ref}}(1 + \varepsilon)$ gives
\[
V_{\text{out}}' = V_{\text{in}} \frac{R_{\text{ref}}(1 + \varepsilon)}
                 {R_{\text{ref}}(2 + \varepsilon)}
              = \frac{V_{\text{in}}}{2} \cdot
                 \frac{1 + \varepsilon}{1 + \varepsilon/2}.
\]
Expanding for small $\varepsilon$,
$\tfrac{1 + \varepsilon}{1 + \varepsilon/2} \approx
1 + \varepsilon - \varepsilon/2 = 1 + \varepsilon/2$, so
$V_{\text{out}}' \approx (V_{\text{in}}/2)(1 + \varepsilon/2)$ and
$\Delta V_{\text{out}}/V_{\text{out}} \approx \varepsilon/2$. The
divider therefore halves the relative resistance perturbation when
read at the symmetric point. The implication for the working
uncertainty budget is that a $\pm 1\,\%$ reference resistor
contributes about $\pm 0.5\,\%$ to the divider's relative output
uncertainty, and the engineer should size the reference resistor
tolerance with the project's per-degree budget in mind.

\begin{exercise}
For a strain gauge with gauge factor $G$, the fractional change in
resistance under axial strain $\varepsilon$ is $\Delta R / R = G
\varepsilon$. Show that, for a $G = 2$ foil gauge under strain
$\varepsilon = 10^{-3}$, the resistance change is approximately
$0.2\,\%$ of the unstrained resistance, and identify the
discrimination required of the readout circuit.
\end{exercise}

\paragraph{Solution sketch.} Direct substitution:
$\Delta R / R = G \varepsilon = 2 \times 10^{-3} = 2 \times 10^{-3}
= 0.2\,\%$. For a $R = 350 \,\si{\ohm}$ foil gauge, this is
$0.7 \,\si{\ohm}$. The readout circuit (typically a Wheatstone
bridge) must discriminate at this level relative to the
unstrained resistance baseline. A single-arm bridge excited at
$V_{\text{ex}} = 5 \,\si{\volt}$ produces an output of
$V_{\text{br}} \approx V_{\text{ex}} \Delta R/(4 R) =
5 \times 0.002/4 = 2.5 \,\si{\milli\volt}$ at full strain; a
$24$-bit ADC at this range gives nominal microvolt-per-LSB
resolution but the practical limit is the bridge thermal
common-mode and the instrumentation amplifier noise floor. The
working production configuration uses a four-arm active bridge
(two gauges in tension, two in compression) that quadruples the
output, eliminates the bridge's thermal common-mode response, and
is the standard load-cell and pressure-transducer architecture.

\begin{exercise}
A piezoelectric accelerometer has charge sensitivity $S_{q}$
(coulombs per metre per second squared). Show that the open-circuit
output voltage of the accelerometer connected to a charge
amplifier with feedback capacitance $C_{f}$ is
$V_{\text{out}} = -S_{q} a / C_{f}$, where $a$ is the input
acceleration. Comment on the high-pass character of the response.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} A piezoelectric crystal produces charge
proportional to applied force; for an accelerometer the proof mass
turns acceleration into force, so the charge output is
$Q = S_{q} a$ with $S_{q}$ in coulombs per metre per second
squared. A \engterm{charge amplifier} is an inverting op-amp
configuration with a capacitor $C_{f}$ across the feedback path
that integrates the input charge into a voltage, plus (in
practice) a high-value resistor $R_{f}$ in parallel that drains
the capacitor at DC and prevents output saturation from leakage
currents.

\textbf{Given.} Sensor charge sensitivity $S_{q}$, input
acceleration $a$, feedback capacitance $C_{f}$, feedback
resistance $R_{f}$. The op-amp is ideal at the working frequency
(infinite gain, infinite input impedance, zero offset).

\textbf{Step 1: charge balance at the summing junction.} With the
non-inverting input grounded and infinite open-loop gain, the
inverting input is held at virtual ground. All sensor charge
$Q = S_{q} a$ must flow into the feedback impedance. For the
capacitor alone, $Q_{C_{f}} = -C_{f} V_{\text{out}}$ (the sign
arises because the inverting topology drives $V_{\text{out}}$
opposite to the input).

\textbf{Step 2: solve for the output voltage.} Equating charges,
$S_{q} a = -C_{f} V_{\text{out}}$, so
$V_{\text{out}} = -S_{q} a / C_{f}$.

\textbf{Step 3: identify the high-pass corner introduced by
$R_{f}$.} At frequency $\omega$ the feedback impedance is the
parallel combination of $1/(j\omega C_{f})$ and $R_{f}$, which
equals $R_{f}/(1 + j\omega R_{f} C_{f})$. The transfer function
from input current $I_{\text{in}} = \dd Q/\dd t = j\omega S_{q}
A$ (with $A$ the phasor amplitude) to $V_{\text{out}}$ is
$V_{\text{out}}/A = -j\omega S_{q} \times R_{f}/(1 + j\omega R_{f}
C_{f})$, a first-order high-pass response with corner $f_{c} = 1/
(2\pi R_{f} C_{f})$ and high-frequency asymptote $S_{q}/C_{f}$.

\textbf{Sanity check.} At a representative laboratory
accelerometer, $C_{f} = 1 \,\si{\nano\farad}$, $R_{f} = 1 \,\si
{\giga\ohm}$ gives $f_{c} = 1/(2\pi \times 10^{9} \times 10^{-9})
\approx 0.16 \,\si{\hertz}$, a fraction of a hertz, consistent
with the chapter's claim. A ruggedised field unit with $C_{f} =
1 \,\si{\nano\farad}$ and $R_{f} = 100 \,\si{\mega\ohm}$ gives
$f_{c} \approx 1.6 \,\si{\hertz}$, several hertz. The trade-off
is that lower $R_{f}$ improves DC stability against bias currents
at the cost of pushing the low-frequency band edge up.

\textbf{Common error.} Treating the piezoelectric accelerometer
as a DC-coupled sensor and trying to measure constant tilt with
it. The signal at $f \ll f_{c}$ rolls off to zero, and the tilt
appears as an exponentially decaying transient. For static or
near-static measurements, the working choice is a capacitive or
piezoresistive MEMS accelerometer with intrinsic DC response.
The reader who must use a piezoelectric unit for low-frequency
work can extend the band only at the cost of larger $C_{f}$ or
$R_{f}$, with their associated stability and noise penalties.

\begin{exercise}
A photodiode produces a current proportional to incident optical
power: $I_{\text{ph}} = R_{\lambda} P$, where $R_{\lambda}$ is the
responsivity at the wavelength of interest. Derive the trans-
impedance amplifier output voltage $V_{\text{out}} = R_{f} I_{\text
{ph}}$ for feedback resistance $R_{f}$, and identify the noise
sources that set the minimum detectable optical power.
\end{exercise}

\paragraph{Solution sketch.} A transimpedance amplifier holds the
photodiode's cathode (or anode, depending on bias polarity) at the
op-amp's virtual ground. The photocurrent $I_{\text{ph}}$ flows
through $R_{f}$, producing
$V_{\text{out}} = R_{f} I_{\text{ph}} = R_{f} R_{\lambda} P$. The
minimum detectable optical power is bounded by three noise sources:
(i) Johnson noise of the feedback resistor,
$v_{n,R} = \sqrt{4 k_{B} T R_{f} \Delta f}$, which translates to an
input-referred current noise
$i_{n,R} = v_{n,R}/R_{f} = \sqrt{4 k_{B} T \Delta f / R_{f}}$;
(ii) shot noise of the photocurrent and any dark current,
$i_{n,\text{sh}} = \sqrt{2 q (I_{\text{ph}} + I_{\text{dk}})
\Delta f}$; (iii) op-amp input current and voltage noise
(translated to a current noise via the photodiode capacitance and
the loop response). The minimum detectable power is then
$P_{\min} \approx i_{n,\text{tot}}/R_{\lambda}$ where
$i_{n,\text{tot}}$ is the quadrature sum of the three. For typical
silicon photodiodes ($R_{\lambda} \sim 0.5 \,\si{\ampere\per\watt}$
at $\sim 800 \,\si{\nano\meter}$) and $R_{f} = 1 \,\si{\mega\ohm}$
over a $1 \,\si{\hertz}$ bandwidth at room temperature, the
Johnson-noise floor gives a minimum detectable optical power of
order $10 \,\si{\femto\watt}$, below which the photodiode is not
the limiting instrument.

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the temperature rise of a thermistor in still air due to
self-heating, assuming the thermistor dissipates $1 \,\si{\milli
\watt}$ and has a thermal resistance to ambient of order $200 \,\si
{\kelvin\per\watt}$. Comment on whether self-heating is a dominant
source of bias in the project's measurement.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} A thermistor dissipates electrical power $P = V
I = V^{2}/R$ when current flows through it. The dissipated power
raises the thermistor's body temperature above ambient by an
amount set by the thermal resistance $\theta_{\text{th}}$ to the
surroundings: $\Delta T = P \theta_{\text{th}}$. Because the
thermistor is also the temperature sensor, the self-heating
raises the inferred temperature above the actual ambient, a
systematic bias that the engineer either avoids by lowering $P$
or compensates by subtracting the modelled self-heating offset.

\textbf{Given.} $P = 1 \,\si{\milli\watt}$, $\theta_{\text{th}} =
200 \,\si{\kelvin\per\watt}$ in still air. Project budget for
expanded uncertainty: $\pm 1 \,\degC$ at $k = 2$.

\textbf{Step 1: compute the steady-state self-heating offset.}
$\Delta T = P \theta_{\text{th}} = 10^{-3} \,\si{\watt} \times 200
\,\si{\kelvin\per\watt} = 0.20 \,\degC$.

\textbf{Step 2: place the offset in the project's uncertainty
budget.} For the $\pm 1 \,\degC$ expanded budget at $k = 2$, the
standard uncertainty target on each component is roughly $0.5 \,
\degC$; if self-heating is treated as a rectangular bias on
$[0, 0.2] \,\degC$ (the offset always raises the reading above
ambient), its standard uncertainty contribution is $0.20/\sqrt{12}
\approx 0.058 \,\degC$, about $12\,\%$ of the per-component
budget. Worth attention; not dominant.

\textbf{Step 3: list the two operative mitigations.} (i)
\emph{Lower the dissipation.} Doubling $R_{\text{ref}}$ in the
divider halves the current at the symmetric operating point and
quarters $P$, dropping the offset to $0.05 \,\degC$. The cost is
higher source impedance at the ADC node and greater susceptibility
to op-amp input bias current. (ii) \emph{Raise the heat-transfer
coefficient.} Immersing the thermistor in a stirred water bath
drops $\theta_{\text{th}}$ by roughly an order of magnitude
(typical $\theta_{\text{th}}$ in stirred water is $10$ to $30
\,\si{\kelvin\per\watt}$), cutting the offset to about $0.02 \,
\degC$. The water bath is the calibration configuration in the
chapter project; the air configuration relies on mitigation (i).

\textbf{Sanity check.} At the divider's symmetric operating point,
the current through the thermistor is $V_{\text{in}}/(2 R_{\text
{ref}}) = 3.3/(2 \times 10{,}000) = 0.165 \,\si{\milli\ampere}$
and the dissipated power is $I^{2} R_{\text{th}} = (1.65 \times
10^{-4})^{2} \times 10{,}000 = 2.7 \times 10^{-4} \,\si{\watt}
\approx 0.3 \,\si{\milli\watt}$, somewhat below the stated $1
\,\si{\milli\watt}$ working envelope. At this lower dissipation
the offset drops to $\Delta T \approx 0.06 \,\degC$, well within
the budget.

\textbf{Common error.} Treating self-heating as a random rather
than systematic effect. Self-heating always raises the reading;
the engineer who averages many samples does not average away the
offset. The mitigation is either to design it out (lower $P$) or
to subtract the modelled offset before reporting; averaging is
the wrong tool.

\begin{exercise}
Estimate the input impedance required of an op-amp voltage follower
to read a thermistor voltage divider without loading it by more
than $1\,\%$. State your assumed source impedance from the divider's
parallel combination.
\end{exercise}

\paragraph{Solution sketch.} The Th\'evenin source impedance seen
at the ADC node is $R_{\text{ref}} \parallel R_{\text{th}}$. At the
symmetric operating point ($R_{\text{ref}} = R_{\text{th}} = 10
\,\si{\kilo\ohm}$) this is $5 \,\si{\kilo\ohm}$. To load by less
than $1\,\%$, the buffer's input impedance must be at least
$100 R_{\text{src}} = 500 \,\si{\kilo\ohm}$. Practical CMOS op-amps
(input bias currents of order picoamperes, input resistance of
gigaohms) exceed this requirement by many orders of magnitude;
bipolar op-amps with nanoampere bias currents are still adequate
for a kilohm source impedance. The constraint becomes operative
only when the source impedance reaches megohms, where bias-current
flow through the source produces a static offset that drifts with
temperature.

\begin{exercise}
Estimate the dominant noise contribution at room temperature in a
$10 \,\si{\kilo\ohm}$ resistor over a $10 \,\si{\hertz}$ measurement
bandwidth, using the Johnson-noise formula
$v_{\text{n}} = \sqrt{4 k_{\text{B}} T R \Delta f}$. State whether
the resulting noise voltage is significant for the thermistor
project's resolution requirement.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} Johnson--Nyquist noise is the open-circuit voltage
fluctuation of a resistor in thermal equilibrium. The mean-square
voltage over a measurement bandwidth $\Delta f$ is
$\langle v_{n}^{2} \rangle = 4 k_{B} T R \Delta f$, with $k_{B}$
the Boltzmann constant, $T$ the absolute temperature, and $R$ the
resistance. The first-principles derivation
\cite{paper:ch05-johnson-1928, paper:ch05-nyquist-1928} from
equipartition was contemporaneous with the formulation of the
fluctuation-dissipation relation.

\textbf{Given.} $R = 10 \,\si{\kilo\ohm}$, $T = 295 \,\si{\kelvin}$
(room), $\Delta f = 10 \,\si{\hertz}$, $k_{B} = 1.381 \times
10^{-23} \,\si{\joule\per\kelvin}$.

\textbf{Step 1: compute the noise spectral density.}
$4 k_{B} T = 4 \times 1.381 \times 10^{-23} \times 295 = 1.630
\times 10^{-20} \,\si{\joule}$. Multiplied by $R$, the
voltage spectral density is $4 k_{B} T R = 1.630 \times 10^{-16}
\,\si{\volt\squared\per\hertz}$, or in
amplitude $\sqrt{4 k_{B} T R} = 1.28 \times 10^{-8} \,\si{\volt}/
\sqrt{\si{\hertz}} = 12.8 \,\si{\nano\volt}/\sqrt{\si{\hertz}}$.
The $12.8 \,\si{\nano\volt}/\sqrt{\si{\hertz}}$ value is the
working figure for a $10 \,\si{\kilo\ohm}$ resistor at room
temperature.

\textbf{Step 2: multiply by the bandwidth.}
$\langle v_{n}^{2} \rangle = 1.630 \times 10^{-16} \times 10 =
1.63 \times 10^{-15} \,\si{\volt\squared}$. Taking the square root,
$v_{n} = \sqrt{1.63 \times 10^{-15}} \approx 4.04 \times 10^{-8}
\,\si{\volt} \approx 40 \,\si{\nano\volt}$ rms.

\textbf{Step 3: compare against the project's signal scale.} One
LSB on the project's $10$-bit ADC at $V_{\text{ref}} = 3.30 \,\si
{\volt}$ is approximately $3.2 \,\si{\milli\volt}$. The
Johnson-noise rms is $40 \,\si{\nano\volt}/0.003226 \,\si{\volt}
\approx 1.2 \times 10^{-5}$ LSB, five orders of magnitude below
the ADC's quantisation step.

\textbf{Sanity check.} Johnson noise scales as $\sqrt{R}$ and as
$\sqrt{\Delta f}$; a $1 \,\si{\mega\ohm}$ resistor at the same
temperature and bandwidth has ten times the rms noise, or $400
\,\si{\nano\volt}$, still well below the project's signal floor.
Johnson noise reaches the LSB only at source impedances around
$10^{12} \,\si{\ohm}$ or bandwidths around $10^{10} \,\si{\hertz}$,
neither of which applies. The bench result in
\repopath{code/johnson_noise.py} confirms the same scaling
empirically.

\textbf{Common error.} Reporting the rms noise without specifying
the bandwidth over which it was integrated. The
$v_{n}/\sqrt{\Delta f}$ spectral density (in volts per root
hertz) is a property of the resistor; the integrated rms voltage
is a property of the resistor and the measurement bandwidth.
Confusing the two leads to noise quotes that disagree by orders
of magnitude depending on what the engineer assumed for the
bandwidth.

\begin{exercise}
Estimate the sample rate at which a smartphone microphone digitises
audio for voice recording, and the resulting Nyquist frequency.
Compare to the working bandwidth of the human voice and identify
the signal-content fraction that the smartphone captures.
\end{exercise}

\paragraph{Solution sketch.} Smartphone voice recording typically
samples at $44.1 \,\si{\kilo\hertz}$ (CD-derived rate) or
$48 \,\si{\kilo\hertz}$ (broadcast-derived rate); the Nyquist
frequency is half this, $22.05$ or $24 \,\si{\kilo\hertz}$. The
voiced human voice has fundamentals from $\sim 80 \,\si{\hertz}$
(low male) to $\sim 400 \,\si{\hertz}$ (high female / child) and
formant content out to $\sim 8 \,\si{\kilo\hertz}$, with sibilant
content extending to $\sim 12$--$15 \,\si{\kilo\hertz}$ in close
recordings. The Nyquist frequency above $20 \,\si{\kilo\hertz}$
exceeds the working voice bandwidth, so the smartphone captures
the full voice signal content the human listener perceives. Telephone-grade speech codecs
(narrowband, $8 \,\si{\kilo\hertz}$ sample rate, $4 \,\si{\kilo
\hertz}$ Nyquist) discard sibilants and most of the upper
formants; smartphone voice memo recordings retain them. Compare
with chapter 6 of this volume on sampling theory; the working
margin between Nyquist and signal bandwidth is what an anti-alias
filter occupies.

\begin{exercise}
Estimate the accumulated orientation uncertainty of a smartphone
gyroscope over a $60 \,\si{\second}$ measurement, assuming a
constant bias uncertainty of $10 \,\si{\degree\per\hour}$ and
angular random walk of $0.1 \,\si{\degree\per\sqrt{\hour}}$.
Compare the result with a tilt measurement that needs
$0.5 \,\si{\degree}$ angular uncertainty for the building-height
project of chapter 4.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} An inertial gyroscope's orientation error
accumulates as the integral of two stochastic terms: a bias
term that grows linearly in elapsed time, and an angular random
walk that grows as the square root of elapsed time. The two are
the working summary of the gyroscope's noise budget at short to
medium time scales. A complementary tilt reference (the
accelerometer, used as a tilt sensor against gravity) bounds the
error at long time scales; this exercise bounds the gyro-only
window in which the orientation estimate is trustworthy.

\textbf{Given.} Bias uncertainty $b = 10 \,\si{\degree\per\hour}$;
angular random walk $\eta = 0.1 \,\si{\degree\per\sqrt{\hour}}$;
elapsed time $t = 60 \,\si{\second}$. Target uncertainty
$0.5 \,\si{\degree}$ for the building-height tilt project.

\textbf{Step 1: convert the time to hours.}
$t = 60/3600 = 0.01667 \,\si{\hour}$.

\textbf{Step 2: bias-term contribution.}
$\Delta \theta_{\text{bias}} = b t = 10 \times 0.01667 = 0.167 \,\si
{\degree}$.

\textbf{Step 3: angular-random-walk contribution.}
$\Delta \theta_{\text{ARW}} = \eta \sqrt{t} = 0.1 \times
\sqrt{0.01667} \approx 0.1 \times 0.129 \approx 0.013 \,\si
{\degree}$.

\textbf{Step 4: combine in quadrature.}
$\Delta \theta = \sqrt{(0.167)^{2} + (0.013)^{2}} \approx 0.167
\,\si{\degree}$. The bias term dominates by a factor of about
$13$; the random walk is negligible at this time
scale.

\textbf{Sanity check.} The crossover time at which the two terms
are equal sits at $t_{c} = (\eta/b)^{2}$ in hours: $t_{c} =
(0.1/10)^{2} = 10^{-4} \,\si{\hour} = 0.36 \,\si{\second}$. For
$t > t_{c}$, the bias dominates; for $t < t_{c}$, the random walk
dominates. Smartphone-grade gyros sit at small $t_{c}$, so any
multi-second integration is in the bias-dominated regime. A
ten-minute integration ($t = 0.167 \,\si{\hour}$) reaches
$\Delta \theta \approx 1.7 \,\si{\degree}$ from bias alone,
exceeding the $0.5 \,\si{\degree}$ budget; smartphone-grade
gyroscopes are working instruments for short measurements only.
The Allan-variance literature \cite{paper:ch05-allan-1966}
formalises the time-scale breakdown.

\textbf{Common error.} Quoting the angular random walk as if it
were the dominant term in all regimes. For laboratory-grade gyros
with low bias drift (a few $\si{\degree\per\hour}$ or less), the
random walk dominates over the full minute-scale integration
budget; for smartphone-grade gyros (ten $\si{\degree\per\hour}$
and up), the bias dominates over almost the entire usable
integration window. The two regimes call for different
mitigations: the random-walk-dominated case benefits from
averaging; the bias-dominated case requires periodic
recalibration against a complementary reference. Mixing the two
recipes hurts both.

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
In a language of the reader's choice, simulate a thermistor voltage
divider with $V_{\text{in}} = 3.3 \,\si{\volt}$,
$R_{\text{ref}} = 10 \,\si{\kilo\ohm}$, and an NTC thermistor with
$\beta = 3950 \,\si{\kelvin}$ and $R_{0} = 10 \,\si{\kilo\ohm}$ at
$T_{0} = 25 \,\si{\degreeCelsius}$. Plot $V_{\text{out}}$ versus
temperature across the range $0$ to $80 \,\si{\degreeCelsius}$.
Identify the temperature at which the divider's sensitivity $\dd
V_{\text{out}} / \dd T$ is largest in absolute value (qualitatively;
no calculus needed).
\end{exercise}

\paragraph{Solution sketch.} The reference implementation is
\repopath{code/thermistor_divider.py}; running the script with
the parameters above tabulates $(T, R_{\text{th}}, V_{\text{out}})$
in $5 \,\degC$ steps and reports the temperature of maximum
$|\dd V/\dd T|$ numerically. The maximum sits near
$T \approx 25 \,\degC$, where the divider is at its symmetric
point ($R_{\text{th}} \approx R_{\text{ref}}$); the slope is
approximately $-0.040 \,\si{\volt\per\kelvin}$ at this point.
The transfer curve in
\Cref{fig:vol01:ch05:thermistor-divider} shows the shape
qualitatively. The reader's own script should match the values to
three significant figures.

\begin{exercise}
Add Gaussian noise to the simulated divider output of the previous
exercise with standard deviation equal to one ADC LSB for a
$10$-bit ADC at the $3.3 \,\si{\volt}$ reference. Simulate $1000$
independent measurements at each of $0$, $25$, $50$, and
$75 \,\si{\degreeCelsius}$. Compute the empirical standard
deviation of the inferred temperature at each point and report the
$T$-dependence of the noise.
\end{exercise}

\paragraph{Solution sketch.} Reference implementation:
\repopath{code/adc_noise_simulation.py}. One LSB is approximately
$3.2 \,\si{\milli\volt}$. The script back-solves through the
divider and the $\beta$-parameter model to recover an inferred
temperature for each noisy sample. Representative results from the
seeded run: $\sigma_{T}(0 \,\degC) \approx 0.05 \,\degC$;
$\sigma_{T}(25 \,\degC) \approx 0.08 \,\degC$;
$\sigma_{T}(50 \,\degC) \approx 0.15 \,\degC$;
$\sigma_{T}(75 \,\degC) \approx 0.40 \,\degC$. The noise grows
with temperature because the divider's sensitivity
$|\dd V/\dd T|$ decreases as the thermistor resistance drops
below $R_{\text{ref}}$; the same ADC noise translates to a larger
temperature uncertainty when each LSB covers more degrees. For the
project's $\pm 1 \,\degC$ budget, ADC noise is acceptable across
the range; for $\pm 0.1 \,\degC$, the engineer either picks a
higher-bit ADC, averages more samples, or chooses an
$R_{\text{ref}}$ that pushes the symmetric point toward the
higher-temperature end.

\begin{exercise}
Fit the Steinhart-Hart equation $1/T = A + B \ln(R/R_{\ast}) +
C [\ln(R/R_{\ast})]^{3}$ to a supplied synthetic calibration table
for one named thermistor part. The problem statement gives the part
number, the source of the nominal coefficients, and the generated
noisy data. Compare the fitted curve with the
$\beta$-parameter approximation over $0$ to $80 \,\si{\degree
Celsius}$ and report the maximum temperature difference.
\end{exercise}

\paragraph{Solution sketch.} Reference implementation:
\repopath{code/steinhart_hart_fit.py}. The fit is linear in
$(A, B, C)$ given $\ln R$ and $1/T$, so the ordinary
least-squares solution closes in one call to
\verb|np.linalg.lstsq|. The synthetic table in the script uses a
$\beta$-model base curve plus a small higher-order perturbation
chosen so that the three-coefficient fit measurably outperforms
the two-coefficient $\beta$ model. Representative results:
maximum $|T_{\text{SH}} - T_{\text{true}}| \approx 0.05 \,\degC$
across $0$--$80 \,\degC$ versus
$|T_{\beta} - T_{\text{true}}| \approx 0.4 \,\degC$ at the
extremes. The Steinhart-Hart fit absorbs the higher-order term;
the $\beta$ model cannot. For a one-degree project the $\beta$
model is adequate near the centre of the calibrated range and
deteriorates at the extremes; for a $\pm 0.1 \,\degC$ project the
Steinhart-Hart form is the working choice. The original paper
\cite{paper:ch05-steinhart-hart-1968} introduced the
log--cubic form in the oceanographic temperature-measurement
context.

\begin{exercise}
Simulate a redundant sensor system with three temperature sensors
each of standard uncertainty $\sigma$, with correlation coefficient
$\rho$ between any pair. Compute the standard uncertainty of the
mean across $\rho \in \{0, 0.5, 0.9, 0.99, 1.0\}$ and plot the
result. Identify the value of $\rho$ at which the redundancy
delivers no benefit relative to a single sensor.
\end{exercise}

\paragraph{Solution sketch.} Reference implementation:
\repopath{code/redundancy_correlation.py}. The closed form is
$u(\bar y)/\sigma = \sqrt{(1 + 2\rho)/3}$, easily verified by
Monte Carlo with multivariate normals. Values:
\begin{center}
\begin{tabular}{rcc}
\toprule
$\rho$ & $u(\bar y)/\sigma$ (closed) & MC \\
\midrule
0.00 & 0.577 & 0.578 \\
0.50 & 0.816 & 0.816 \\
0.90 & 0.966 & 0.966 \\
0.99 & 0.996 & 0.997 \\
1.00 & 1.000 & 1.000 \\
\bottomrule
\end{tabular}
\end{center}
At $\rho = 1$ the redundancy delivers no benefit; at $\rho = 0.9$
the gain is small (about $3\,\%$ reduction); at $\rho = 0$ the
gain is the textbook $\sqrt{3}$. The plot in
\Cref{fig:vol01:ch05:common-mode} shows the curve. The AF447
pitot-icing event sits at $\rho \to 1$ for the duration of the
common environmental cause.

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a sensor selection for a clinical-grade thermometer measuring
human core body temperature ($35$ to $42 \,\si{\degreeCelsius}$)
with required expanded uncertainty $\pm 0.1 \,\si{\degreeCelsius}$
at $k = 2$. State the sensing principle chosen, the rationale, the
dominant uncertainty source you expect, and the calibration
procedure.
\end{exercise}

\paragraph{Discussion.} A defensible answer identifies a calibrated
NTC thermistor probe or a class-A Pt100 RTD as the sensing element
(both reach the budget); rejects type-K thermocouples on
sensitivity grounds (the $40 \,\si{\micro\volt\per\kelvin}$ Seebeck
coefficient is hard to read to a sufficient fraction of a degree
within the budget); identifies the calibration uncertainty against
a traceable reference as the dominant a posteriori source (the
thermistor's intrinsic resolution is well below the budget); and
specifies a calibration at three or more reference points across
the body-temperature range using a NIST-traceable reference
thermometer in a stirred bath, with an explicit recalibration
interval (per the second estimation block in section 5.4). A
strong answer also identifies the regulatory regime (IEC 60601-2-56
for clinical thermometers; the Medical Devices Regulation in the
EU; FDA 510(k) in the US) and accounts for it in the design and
documentation burden; a weaker answer treats the device as if it
were an industrial instrument. The Hubble lesson applies in
miniature: the reference thermometer must itself be checked, ideally
against an ITS-90 fixed-point cell or a calibration laboratory's
working standard, not against another reference of the same type.

\begin{exercise}
Design a measurement chain to record the vibration of a household
appliance (refrigerator compressor, washing machine, dishwasher) for
diagnostic purposes. Specify the sensor (accelerometer
specifications), the signal conditioning, the sample rate, the
data-acquisition duration, and the analysis. State the failure
modes the measurement could distinguish.
\end{exercise}

\paragraph{Discussion.} A working answer chooses a piezoelectric
or MEMS accelerometer with a working bandwidth of at least
$500 \,\si{\hertz}$ (to capture compressor harmonics) and a full
scale of order $\pm 10 \,\si{\meter\per\second\squared}$ (a few
$g$); specifies a sample rate of at least $5 \,\si{\kilo\hertz}$
(a Nyquist of $2.5 \,\si{\kilo\hertz}$ with comfortable margin
above the highest signal frequency of interest, anti-alias
filtered); records for a duration that spans multiple operating
cycles (one to ten minutes for a compressor); and specifies
post-processing as the power spectral density or short-time
Fourier transform of the acceleration signal. The failure modes
the measurement can distinguish include unbalanced rotating mass
(harmonics of the rotational frequency growing over time), bearing
wear (broadband noise above the fundamental), loose mounting (low-
frequency rocking modes), and structural resonance (a stationary
peak that appears when a particular operating mode is engaged).
A weaker answer either over-specifies (industrial accelerometer
for kitchen appliance) or under-specifies (smartphone accelerometer
with $50 \,\si{\hertz}$ effective bandwidth, missing the working
diagnostic content). The IEPE interface
\cite{text:ch05-pallas-areny-webster2001} is the working production
choice for piezoelectric units.

\begin{exercise}
Design a sensor configuration for a robotic manipulator's end-
effector position with required combined standard uncertainty
$\pm 0.1 \,\si{\milli\meter}$ over a $300 \,\si{\milli\meter}$
working range. State the sensing principle chosen, the
specifications required of each sensor, and the data fusion
strategy if multiple sensors are used.
\end{exercise}

\paragraph{Discussion.} A working answer either reads each joint
angle with a high-resolution incremental or absolute encoder
(typically optical, $\sim 0.001 \,\si{\degree}$ resolution; combined
with the manipulator kinematic model, the end-effector position
uncertainty is set by encoder resolution plus joint backlash plus
link compliance) or measures end-effector position directly with a
laser tracker or a precision LVDT array. Direct end-effector
measurement avoids the kinematic-error stack but introduces line-of-
sight and mounting constraints. Strong answers identify the trade-
off explicitly: encoder-only is cheap and well-precedented but accumulates
error through the kinematic chain; direct measurement closes the
loop on what the engineer actually cares about (end-effector
position) at the cost of an extra sensor and its mounting. Data
fusion would typically combine the high-rate encoder measurement
with a slower, more accurate direct measurement via a Kalman
filter or a complementary filter; the encoder provides smooth
high-frequency motion, the direct sensor calibrates the long-term
bias. Calibration of the kinematic model itself (using a
calibrated artefact moved through the workspace) is the working
mitigation for kinematic-chain error stacking.

\begin{exercise}
Design a sensor selection for measuring atmospheric carbon dioxide
concentration indoors with required range $400$ to
$5{,}000\,\text{ppm}$ and combined standard uncertainty
$\pm 50\,\text{ppm}$ at $k = 2$. State the sensing principle, the
expected interferences, the calibration procedure, and the
recalibration interval.
\end{exercise}

\paragraph{Discussion.} A defensible answer picks a non-dispersive
infrared (NDIR) CO$_{2}$ sensor: an IR source, an optical filter
centred at the $4.26 \,\si{\micro\meter}$ CO$_{2}$ absorption
line, a thermopile detector, and the Beer--Lambert inversion to
recover concentration. Working-grade NDIR units reach the budget
across the indoor range. Expected interferences include humidity
(broadband absorption, partly compensated by the optical filter
and partly by an auxiliary humidity sensor), temperature (the
detector and the source drift with ambient), and altitude (the
inversion assumes a known cell pressure). The calibration
procedure exposes the sensor to two known concentrations (typically
outdoor ambient $\sim 420 \,\text{ppm}$ as of 2026, and a bottled
reference of $1{,}000$ or $2{,}000 \,\text{ppm}$) and fits the
Beer--Lambert inversion. Indoor working units typically include
an automatic baseline-correction algorithm that assumes the
sensor sees the outdoor ambient every few days and recalibrates
the offset; the algorithm fails in continuously occupied spaces
that never reach $420 \,\text{ppm}$. The recalibration interval
for manual calibration is typically six to twelve months for
working indoor units; a strong answer notes that this is the
manufacturer's interval and that the project's actual interval
depends on its drift budget per the second estimation block.
Electrochemical CO$_{2}$ sensors are an alternative but generally
do not reach the project's uncertainty budget.

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A thermocouple-based temperature measurement reads $25 \,\si{\degree
Celsius}$ steadily but the reference thermometer reads $30 \,\si
{\degreeCelsius}$. State three plausible mechanisms (cold-junction
compensation error, thermocouple wire substitution, calibration
drift, thermal contact issue) and propose the next measurement that
distinguishes them.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} A thermocouple reads $5 \,\degC$ low against a
trusted reference. The mismatch is a single observation; the
diagnostic question is which of four working mechanisms produced
it. The four candidates (cold-junction compensation error, wire
alloy substitution, calibration drift, thermal contact) cover the
dominant failure modes of a thermocouple measurement chain. The
diagnostic strategy is to test each in order of cost, ruling
candidates in or out with a single bench operation each.

\textbf{Given.} Thermocouple indicates $25 \,\degC$; reference
thermometer indicates $30 \,\degC$; offset $-5 \,\degC$;
nominally a type-K junction with internal cold-junction
compensation.

\textbf{Step 1: cold-junction compensation error.} The internal
cold-junction sensor may read the wrong reference temperature; if
it reads $5 \,\degC$ high, the inferred measuring-junction
temperature is $5 \,\degC$ low.
\emph{Test}: move the thermocouple measuring junction into an ice
bath; the indicated temperature should read $0 \,\degC$ within
the working accuracy if the compensation is correct, $5 \,\degC$
if the compensation is the cause.

\textbf{Step 2: wrong wire alloy.} An extension wire of the wrong
alloy (e.g., type-J wire substituted for type-K) has a different
Seebeck coefficient; the instrument's polynomial decoder returns
a wrong-by-coefficient-ratio temperature.
\emph{Test}: replace the lead wire with a known-good type-K
extension and reread; if the offset disappears, wire alloy was
the cause. The fingerprint of this failure mode is a
proportional, not constant, offset that grows with $\Delta T$.

\textbf{Step 3: calibration drift.} The thermocouple's
sensitivity has drifted from its calibrated value through aging,
oxidation, or sustained high-temperature exposure.
\emph{Test}: recalibrate against the reference thermometer at
two or three temperatures across the working range; a
calibration drift typically appears as a slope shift, not just
an offset.

\textbf{Step 4: thermal contact issue.} The measuring junction is
not at the reference thermometer's temperature: the two probes
are in different media (one in air, one in water; one in fast-
flowing fluid, one in stagnant fluid; one with good thermal
contact, one without).
\emph{Test}: bring the two probes into direct mechanical contact
and verify they read the same; any persistent offset is then in
the sensing chain, not the thermal coupling.

\textbf{Sanity check.} The ice-bath test in Step 1 is the most
direct (and cheapest); running it first either confirms the
compensation as the cause or removes it from the candidate list
with a single recording. The wire-alloy test requires a spare
known-good wire and a few minutes of swap time; the recalibration
test is the most expensive but the only one that distinguishes
slope from offset; the thermal-contact test is a procedural
check, not a hardware change. The four-step order minimises
total diagnostic cost.

\textbf{Common error.} Diagnosing the failure by inspection
without a test. The four mechanisms have superficially similar
signatures (constant offset); only the targeted test
distinguishes them. The reader who guesses "calibration drift"
because the sensor is old, replaces it, and finds the offset
unchanged has discovered that calibration drift was not the
cause and consumed a sensor in the process. The discipline is to
run the tests; the discipline is what separates a working
diagnostic process from guesswork.

\begin{exercise}
A pressure transducer's output drifts $0.05\,\%$ of full-scale per
hour during a long test. State three plausible drift mechanisms
(temperature variation, humidity ingress, electronic component
warm-up, mechanical relaxation) and the diagnostic test for each.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} A $0.05\,\%$-of-full-scale drift per hour over a
long test is the residue of one or several time-varying physical
mechanisms acting on the transducer's mechanical, electrical, or
chemical interfaces. The diagnostic strategy is to vary one
candidate cause at a time and observe whether the drift signature
changes. The four named mechanisms (temperature variation,
electronic warm-up, mechanical relaxation, humidity ingress) each
leave a different signature in time, in ambient correlation, and
in mechanical history.

\textbf{Given.} A pressure transducer's output drifts $0.05\,\%$
of full scale per hour during a long test under nominally constant
applied pressure. Working hypothesis space: the four named
mechanisms. Goal: distinguish them with the smallest test sequence.

\textbf{Step 1: temperature variation.} The transducer's
temperature coefficient is order $0.01\,\%/\degC$; a $5 \,\degC$
ambient drift over a few hours explains the observation. The
signature is correlation between drift and ambient temperature,
visible as a scatter plot of output versus temperature with a
clear slope.
\emph{Test}: log ambient temperature alongside the transducer
output; compute the correlation coefficient and the slope. A
correlation coefficient $|r| > 0.7$ and a slope consistent with
the manufacturer's spec is enough to assign the drift to
temperature.

\textbf{Step 2: electronic warm-up.} The transducer's amplifier
and reference-voltage source warm up over the first one to two
hours of operation. The signature is an exponentially decaying
drift with time constant of order one hour, independent of ambient
temperature, that asymptotes once the internal thermal mass
stabilises.
\emph{Test}: leave the transducer powered for several hours before
the test begins, then start the recording; if the drift signature
has disappeared, electronic warm-up was the cause.

\textbf{Step 3: mechanical relaxation.} A diaphragm or its mounting
relaxes under sustained pressure load; the elastic-to-plastic
boundary slowly shifts. The signature is a drift that grows
quasi-logarithmically in time, depends on the magnitude of the
sustained pressure, and is partly reversible by cycling.
\emph{Test}: cycle the transducer between zero and full-scale
several times before the test; if the post-cycling drift is
smaller than the original drift, mechanical relaxation was
contributing.

\textbf{Step 4: humidity ingress.} The transducer's seal degrades,
internal humidity rises, and the strain gauge or piezoresistive
element drifts as the bonding adhesive absorbs water. The
signature is a slow drift correlated with ambient humidity, often
non-monotonic over a day as the indoor humidity cycles.
\emph{Test}: monitor internal humidity (if a sensor is present)
or inspect the seal; if humidity ingress is confirmed by visible
moisture or by hygrometer at the connector, the transducer
should be replaced.

\textbf{Sanity check.} The order in which the four tests are run
matters: cheap and fast tests come first. Logging ambient
temperature requires only a second sensor and costs almost
nothing; powering up several hours early is a procedural change;
cycling the transducer is a mechanical operation; opening the
seal is destructive. The diagnostic discipline is to escalate
only when the cheaper tests have ruled their candidates out. The
multi-mechanism case (two simultaneous drifts) is diagnosable
only with multi-factor variation, which is several times more
expensive than the single-cause sequence and is reserved for cases
where the single-cause tests return ambiguous answers.

\textbf{Common error.} Treating drift as a single phenomenon
characterised by a single number. Drift signatures encode
mechanism; the engineer who reports only the rate has discarded
the diagnostic content. The working report for a drifting sensor
records both the magnitude and the shape of the drift over time,
plus the correlated ambient channels (temperature, humidity, line
voltage), so that the mechanism can be inferred from a single
recording rather than requiring a return to the bench.

\begin{exercise}
A strain-gauge-based load cell reads correctly under static load
but produces an apparent reading offset under cyclic loading at
$10 \,\si{\hertz}$. State plausible mechanisms (gauge bond fatigue,
hysteresis, structural dynamics, signal conditioning bandwidth
limit) and propose the test sequence that distinguishes them.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} The load cell reads correctly under static load
and produces an offset under cyclic load at $10 \,\si{\hertz}$.
The offset under dynamic load with static-load agreement is the
diagnostic signature: any mechanism that is symptomatic in steady
state is ruled out; the residual is something that activates only
in the time-varying regime. The four candidates (bandwidth limit,
structural dynamics, hysteresis, bond fatigue) cover the dominant
working classes.

\textbf{Given.} A foil strain-gauge load cell with bridge
excitation, instrumentation amplifier, and digital acquisition.
Static-load reading correct within calibration. Cyclic-load
reading at $10 \,\si{\hertz}$ shows an apparent offset (a
non-zero mean over many cycles).

\textbf{Step 1: bandwidth limit.} If the amplifier or
anti-aliasing filter rolls off below $10 \,\si{\hertz}$, the
acquisition under-samples the peaks of the cyclic waveform; a
half-wave-rectified or averaged reading would then show a bias.
\emph{Test}: drive a known oscillating load at progressively
lower frequencies, say $1$, $3$, $10$, $30 \,\si{\hertz}$. If
the offset is present at $10 \,\si{\hertz}$ and disappears
below the rolloff frequency, the bandwidth is the cause. The
fix is a higher-bandwidth amplifier or a properly anti-aliased
ADC chain.

\textbf{Step 2: structural dynamics.} If the load cell, its
fixture, or the test article has a structural resonance near
$10 \,\si{\hertz}$, the reading reflects the resonance
amplification rather than the applied force. The signature is a
load-cell output that is consistent in magnitude but inconsistent
with the input force amplitude.
\emph{Test}: sweep the drive frequency from $5$ to $15 \,\si
{\hertz}$ in $1 \,\si{\hertz}$ steps; the offset should peak at
the structural-resonance frequency. The fix is to stiffen the
fixture, add damping, or change the drive frequency away from
the resonance.

\textbf{Step 3: hysteresis.} Foil strain gauges and their bonding
adhesives show small elastic hysteresis under cyclic loading; the
rising half of the cycle traces a different path from the falling
half. If the acquisition averages across the asymmetric cycle,
the mean reading shifts.
\emph{Test}: compute the mean from the rising and falling halves
of the cycle separately, $\bar y_{\uparrow}$ and $\bar y_{\downarrow}$;
the half-cycle offset is the hysteresis amplitude, and the
average $(\bar y_{\uparrow} + \bar y_{\downarrow})/2$ removes it.
The fix is either to compensate in software or to choose a
load-cell topology with smaller hysteresis (a Bridge with
two-active-arm topology is half as hysteretic as a single-arm
bridge).

\textbf{Step 4: bond fatigue.} Repeated strain cycling above the
adhesive's working envelope fatigues the gauge-to-substrate bond;
the gauge factor drifts and the calibration shifts.
\emph{Test}: replace the gauge with a fresh part of the same
specification and rerun; if the offset disappears, bond fatigue
was the cause. The fix is either a higher-cycle-life adhesive
or a periodic recalibration schedule that catches the drift
before it accumulates.

\textbf{Sanity check.} The four tests are ordered by cost: the
bandwidth test requires only a different drive frequency; the
resonance sweep requires a longer test sequence; the half-cycle
analysis requires only post-processing of the existing data; the
gauge replacement requires a hardware change. Running them in
that order resolves the question at minimum total cost. A second
sanity check: at the working strain levels for a load cell
(typically below $10^{-3}$), bond fatigue is unlikely after a
short test, so the cheapest tests are also the most likely to
return a positive answer.

\textbf{Common error.} Trying gauge replacement first because it
is the most familiar hardware action. The bandwidth test is faster
and cheaper, and it answers the more common cause; jumping to a
hardware change skips the diagnostic discipline and consumes
both time and a calibrated spare.

\begin{exercise}
A multimeter reads $5.32 \,\si{\volt}$ correctly on its
$\pm 6 \,\si{\volt}$ range but reads $5.31 \,\si{\volt}$ on its
$\pm 60 \,\si{\volt}$ range. State plausible mechanisms (range
calibration, ADC integral nonlinearity, thermal EMF in switching)
and identify which is consistent with the symptom.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} A handheld DMM reads $5.32 \,\si{\volt}$
correctly on its $\pm 6 \,\si{\volt}$ range and $5.31 \,\si
{\volt}$ on its $\pm 60 \,\si{\volt}$ range, a $10 \,\si{\milli
\volt}$ discrepancy that is stable and reproducible. The three
candidate mechanisms (range calibration, ADC integral
nonlinearity, thermal EMF in range switching) have different
quantitative signatures; the diagnostic is to match the symptom
against the signature.

\textbf{Given.} Reading on $6 \,\si{\volt}$ range: $5.32 \,\si
{\volt}$; reading on $60 \,\si{\volt}$ range: $5.31 \,\si{\volt}$;
discrepancy $10 \,\si{\milli\volt}$.

\textbf{Step 1: convert the discrepancy to relative figures.}
$10 \,\si{\milli\volt}$ is $0.17\,\%$ of the $6 \,\si{\volt}$
range reading, $0.02\,\%$ of the $60 \,\si{\volt}$ range
full scale, and $0.19\,\%$ of the reading.

\textbf{Step 2: range-calibration check.} The two ranges have
independent gain-calibration constants stored in the DMM's
non-volatile memory. A $0.02\,\%$ offset between them is well
within the working accuracy specification of a four-and-a-half-
digit handheld DMM (typically $\pm 0.5\,\%$ of reading on each
range), and a stable, reproducible offset between ranges is the
canonical signature of a range-gain mismatch. Most consistent
with the symptom.

\textbf{Step 3: ADC integral nonlinearity check.} The ADC reads
different fractions of full scale on the two ranges: about
$0.886$ FS on the $6 \,\si{\volt}$ range, about $0.088$ FS on
the $60 \,\si{\volt}$ range. The ADC's integral nonlinearity
(INL) curves through both points; if the INL is of order
$0.01\,\%$ FS, the two range readings can differ by a few
millivolts even with perfect range calibration. Also consistent
with the symptom but typically a smaller contribution.

\textbf{Step 4: thermal-EMF check.} A relay or rotary-switch
contact in the range switch can have a thermal EMF of order
$\sim 1 \,\si{\micro\volt}$ on a CuNi-thermocouple-type contact,
swelling to $10 \,\si{\micro\volt}$ on a tin-solder junction.
These are several orders of magnitude smaller than the $10 \,\si
{\milli\volt}$ discrepancy and cannot dominate unless the contact
is defective. Unlikely to be the cause.

\textbf{Sanity check.} The discrepancy is stable and reproducible
across multiple measurements at the same input. A range
calibration mismatch is stable by definition; an INL contribution
is stable for a given input voltage; a thermal-EMF effect would
vary with the contact's temperature, which the symptom does not
report. The two consistent candidates can be distinguished by
measuring at several voltages: a range calibration produces a
linear (proportional) discrepancy; INL produces a curved one.

\textbf{Common error.} Treating any discrepancy between ranges
as a defect. Range calibration mismatch of $0.02\,\%$ FS is
within the working specification of a four-and-a-half-digit DMM;
calling the meter broken would discard a working instrument. The
working diagnosis is: confirm the mismatch is within the meter's
specification at each range; if so, accept the reading on the
lower range as more accurate (more digits used) and report the
range used; if not, calibrate against a precision voltage
reference and adjust the range gain constant on the offending
range per the DMM's service manual.

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
The Air France 447 case is canonical for common-mode sensor failure.
Argue, in 200 to 300 words, what additional sensor diversity (in
sensing principle, in physical location, in calibration history)
would have meaningfully reduced the probability of common-mode
pitot failure in cruise icing conditions, citing the BEA report
\cite{acc:bea-af447}.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} The Air France 447 cruise sequence (Atlantic
crossing, 1 June 2009) lost three independently mounted pitot
probes to common-cause icing within roughly one minute. The
question is not whether the probes were redundant; it is whether
redundancy of the type deployed could be expected to survive the
shared environment that took them all down together. The
diagnostic frame is the three diversity dimensions: sensing
principle, physical location, and supply/calibration history.

\textbf{Given.} Three Thales AA pitot probes, identical model,
mounted on the same nose section, sharing the same heating
control, the same atmospheric environment, and the same
calibration cohort. BEA accident report
\cite{acc:bea-af447} as the primary source of operational and
mechanical detail.

\textbf{Step 1: sensing-principle diversity.} A pitot probe
measures dynamic pressure (the kinetic-energy contribution to
total pressure above static); icing the probe orifice destroys
all three readings together because the failure mechanism is
specific to dynamic-pressure sensing in supercooled cloud. A
complementary measurement that does not depend on the pitot
principle removes this common-mode pathway. The working
sensing-principle alternatives are: synthetic airspeed computed
from inertial-reference acceleration corrected by an atmospheric
model, GPS ground speed corrected by wind estimates from prior
position fixes, and angle-of-attack-derived airspeed from a
calibrated alpha vane. Modern A350 and 787 aircraft carry
synthetic airspeed for exactly this case, as a backup that is
geographically independent of the pitot-icing mechanism.

\textbf{Step 2: physical-location diversity.} The BEA report
notes that the three pitot probes on F-GZCP were on the same
nose section of the fuselage, sharing the same ice environment
during the cruise penetration of the convective system. A probe
mounted at a geometrically distinct position (for example aft of
the wing root, where the local airflow and surface temperature
differ) breaks the geometric common-mode. The engineering cost
is non-trivial: the asymmetric mounting introduces calibration
challenges (different position-error corrections at different
angles of attack), and the airworthiness certification basis for
asymmetric pitot location is less established than the symmetric
case. The trade-off is between the engineering cost of asymmetric
calibration and the risk premium of common-mode icing.

\textbf{Step 3: calibration-history diversity.} All three probes
on F-GZCP came from the same Thales AA batch, with the same
heating-control firmware, the same calibration coefficient. A
fleet-level mitigation against batch-correlated failure modes
would source at least one probe from a different vendor (the
A330 fleet's migration toward Goodrich BA probes was already
underway at the time of the accident, but F-GZCP had not yet
been upgraded \cite{acc:bea-af447}). The mitigation is
structurally similar to the cross-vendor redundancy used in
critical control systems: shared firmware is a common-cause
failure mode even when the hardware is nominally independent.

\textbf{Sanity check.} The three dimensions are necessary, not
sufficient. Sensing-principle diversity alone (a synthetic
airspeed channel) without the crew training to recognise and use
it leaves the same gap; the BEA report places significant weight
on the crew's loss of situational awareness once the primary
airspeed channels failed. The post-AF447 corrective actions
combined synthetic airspeed with revised stall-recovery training
and revised unreliable-airspeed procedures, not just hardware
upgrades.

\textbf{Common error.} Conflating triple-channel redundancy with
diversity. Three identical channels protect against single-
channel hardware failure but contribute nothing against shared
environmental or shared software failure. The diversity framework
must be applied across the four dimensions, environment, geometry,
supply chain, and software, before redundancy is credited against
common-cause events. The same lesson appears in the substation
voltage-transducer exercise below, in a different domain.

\begin{exercise}
A redundant sensor configuration of three voltage transducers is
deployed in a substation, all from the same manufacturer's batch
and calibrated against the same secondary reference. Argue, in
200 to 300 words, the failure modes that this redundancy does and
does not protect against, and identify the additional sensor
diversity (different sensing principle, different manufacturing
batch, different calibration path) that would close each gap.
\end{exercise}

\paragraph{Worked solution.}
\textbf{Set-up.} A triply-redundant sensor configuration is
canonical in power-system instrumentation: three voltage
transducers feed a majority vote, and the substation accepts the
voted value as the working measurement. The question is what
common-cause modes survive the vote, and what additional
diversity closes each.

\textbf{Given.} Three voltage transducers, same manufacturer,
same batch, same secondary reference, in the same rack at the same
substation.

\textbf{Step 1: what the redundancy protects against.}
(i) \emph{Single-channel hardware failure.} One transducer dies;
the other two outvote it; the substation continues operating on
the correct vote. (ii) \emph{Random reading-level noise.} The
mean of three independent zero-mean noise streams has standard
deviation reduced by $\sqrt{3}$.

\textbf{Step 2: batch-correlated manufacturing defect.} All three
transducers share a wafer or assembly run with the same
thermal-coefficient or piezoresistive weakness. Under the same
adverse condition (high ambient temperature, voltage transient,
or aging cycle), all three fail together. The vote does not catch
the failure because all three readings agree on the wrong value.
\emph{Gap closer}: source the three transducers from at least two
different manufacturing batches, ideally two different vendors.

\textbf{Step 3: shared calibration history.} All three transducers
were referenced to the same secondary standard during their last
calibration cycle. If that standard has drifted, all three drift
in the same direction by the same amount, and the vote returns a
unanimous wrong answer.
\emph{Gap closer}: calibrate at least one transducer against an
independent secondary (preferably against a primary). The
recalibration interval should be shorter than the budgeted drift
of either standard.

\textbf{Step 4: shared environment.} All three transducers sit in
the same rack and experience the same temperature, humidity,
vibration, and EMI. Any environment-dependent error appears in
all three readings.
\emph{Gap closer}: physical separation across the substation,
diverse housings, or active environmental compensation that uses
the rack temperature as an input.

\textbf{Step 5: shared sensing principle.} If all three are
voltage-divider-based potential transformers, a saturation event
on the substation primary side saturates all three identically.
\emph{Gap closer}: include at least one alternative sensing
principle (optical voltage transducer, capacitive voltage
divider, or resistive divider with isolated readout), even at
reduced precision, to serve as a sanity check during transient
events.

\textbf{Sanity check.} The four-dimension framework
(batch, calibration, environment, sensing principle) recovers the
same four classes identified in the AF447 worked solution, in a
different domain. The structural pattern, redundant channels
sharing a hidden common cause, is the load-bearing failure mode
of multi-sensor systems. Mapping each new substation deployment
through the four-dimension table is the working defence.

\textbf{Common error.} Claiming high availability from a
nominally redundant configuration without auditing the shared
batch, calibration, environment, and sensing-principle paths.
The vote against single-channel failure is necessary, not
sufficient; the AF447 case and the substation case are the same
pattern in two industries, and the discipline of auditing the
shared causes is the same in both.

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A senior colleague tells the reader that "you only need a sensor
spec sheet, not a calibration certificate, for working
laboratory measurements." Write a one-paragraph response that takes
the colleague seriously, identifies the regime in which the claim
is defensible, and identifies the regime in which it is not.
\end{exercise}

\paragraph{Discussion.} The defensible regime: for relative
measurements (the second reading minus the first, with the same
sensor, within minutes, at the same temperature, on a sensor whose
specification is at least ten times tighter than the required
uncertainty), the spec sheet plus a sanity check at a single
reference point are often adequate. The indefensible regime:
absolute measurements where the result will be compared with
another laboratory's, with a regulatory standard, or with a
historical record; measurements at the edge of the sensor's range
or outside its calibration regime; measurements that depend on the
sensor's specification at a working point distant from where the
manufacturer tested it; and any measurement whose consequence of
error is material. A strong response distinguishes the two regimes
by the working test in chapter 3: a measurement that travels
(across labs, across time, across regulatory bodies) needs the
calibration certificate; a measurement that does not travel may
not. The colleague's claim is half-right under specific
conditions and dangerous when those conditions are not met; the
test in any disputed case is whether the measurement is intended to
travel.

\begin{exercise}
Argue, in 200 to 300 words, whether a smartphone is or is not an
acceptable instrument for engineering-grade measurements. Take a
position; defend it with at least one specific class of measurement
where the smartphone is acceptable and one where it is not.
\end{exercise}

\paragraph{Discussion.} A defensible position: the smartphone is
an acceptable instrument for measurements whose required
uncertainty matches the smartphone's working uncertainty in the
chosen sensor, whose calibration history can be established or is
not load-bearing, and whose result will be used internally rather
than across institutional boundaries. \emph{Acceptable class}:
estimating the inclination of a stair tread using the accelerometer
as a tilt sensor; the measurement budget of a few degrees matches
the smartphone accelerometer's working uncertainty after a single
zero-calibration on a level surface, and the result is internal to
the project. \emph{Unacceptable class}: a regulatory noise survey
of a residential property line for a workplace-safety filing; the
smartphone microphone's frequency response, A-weighting, and
absolute calibration are not traceable to a class 1 sound-level
meter standard, and the regulatory body will reject the
measurement on procedural grounds even if the recorded numbers
were correct. A strong response identifies that the smartphone is
not one instrument but a sensor array (section 5.5) and that the
answer differs by sensor: GPS for navigation-scale position is
acceptable, GPS for centimetre-grade survey is not; the
magnetometer for orientation in the absence of nearby ferrous
mass is acceptable, the magnetometer for absolute magnetic-field
metrology is not. The general principle: the smartphone is
acceptable when its specifications meet the required uncertainty
and its calibration discipline meets the project's traceability
requirement, and not otherwise.
