\chapter{Integration}
\label{vol02:ch06}

\epigraph{Calculus required continuity, and continuity was supposed to
require the infinitely little; but nobody could discover what the
infinitely little might be.}{Bertrand Russell, \emph{The Principles of
Mathematics}, 1903, ch.~XLI.}

\chapmeta{Half-life: foundational. Archetypes: balance (the definite
integral as accounting), transport (the integral of a flux as the total
transported quantity). Exercise target: 40. Project track: household +
analysis (hybrid). Hazard class: none. Reader path default: standard,
with sections explicitly tagged. Named cases: none.}

\noindent
The integral is the engineer's accounting tool. A current that flows
for a measured time leaves a definite charge; a force that acts over a
measured distance leaves a definite work; a probability density taken
over a measured interval leaves a definite probability. Each of these
quantities is the limit of a sum, and that limit is the definite
integral. Chapter 5 introduced the derivative as the formal handling
of rates. This chapter introduces the integral as the formal handling
of accumulation, derives the relation between the two, and equips the
reader with the techniques (analytical, numerical, multivariable, and
convolutional) that engineering work demands.

\section{The integral as accumulation}\pathtag{core}

A reader who has finished Volume I has already integrated by hand. The
total mass of water that flows through a kitchen tap in five minutes
is the flow rate (in litres per minute) times the duration (in
minutes), as long as the rate is constant. When the rate varies in
time, the same accounting must still be done, and the question of how
to do it is the question this chapter answers.

\subsection{The Riemann sum}

A function $f$ is given on the interval $[a,b]$. The reader wants the
total accumulation of $f$ over that interval. The first step is to
partition the interval. Choose points
\[
a = x_{0} < x_{1} < x_{2} < \cdots < x_{n-1} < x_{n} = b,
\]
and on each subinterval $[x_{i-1}, x_{i}]$ pick a sample point
$\xi_{i}$. The \engterm{Riemann sum} is
\[
S_{n} = \sum_{i=1}^{n} f(\xi_{i})\,(x_{i} - x_{i-1}).
\]
Each term is a rate times a width, with the dimensions of an
accumulation. A flow rate in litres per minute, sampled at $\xi_{i}$
and multiplied by the duration $x_{i}-x_{i-1}$ in minutes, gives
litres. The sum has the dimensions of a total volume of water. The
Riemann sum is dimensional accounting before it is mathematics.
\Cref{fig:vol02:ch06:riemann-sum} shows the picture for a
sinusoidally varying integrand: the rectangles overshoot on rising
portions of $f$ and undershoot on falling portions, and the picture
makes the limit-of-mesh argument visible.

\input{volumes/02-form/ch06-integration/figures/fig-riemann-sum.tex}

\begin{archetype}[Balance, accumulation form]
The Riemann sum is the chapter's pure form of the balance archetype
in its accounting variant. A quantity is what it is at each instant;
the total is the sum over all instants, weighted by their share of
the interval. A balance is an accounting equation in space; an
integral is an accounting equation in a continuous variable. The
mass of water that flows through the tap, the charge that crosses a
wire, the heat that enters a room, the probability of an outcome,
the moment of a force, the area under a curve: each is the same
shape of accounting, applied to a different quantity.
\end{archetype}

The width $\Delta x_{i} = x_{i} - x_{i-1}$ is the share of the
interval that the sample $\xi_{i}$ represents. The mesh of the
partition is $\norm{P} = \max_{i} \Delta x_{i}$. As the mesh shrinks
toward zero, three things happen at once. The number of subintervals
grows; the share each one represents shrinks; the chosen sample
matters less because the function has less room to change inside the
subinterval. When the function is well-enough behaved on the
interval, all sequences of partitions whose mesh tends to zero, with
all choices of sample points, give the same limit. That common limit
is the integral.

\subsection{The definite integral}

\begin{definition}[Riemann integral]
A bounded function $f \colon [a,b] \to \R$ is \engterm{Riemann
integrable} on $[a,b]$ if there exists a number $I$ such that for
every $\varepsilon > 0$ there is a $\delta > 0$ with
\[
\abs*{\,\sum_{i=1}^{n} f(\xi_{i})\,\Delta x_{i} - I\,} < \varepsilon
\]
for every partition $P$ of $[a,b]$ with $\norm{P} < \delta$ and
every choice of sample points $\xi_{i} \in [x_{i-1}, x_{i}]$. The
number $I$ is the \engterm{definite integral} of $f$ over $[a,b]$
and is written
\[
\int_{a}^{b} f(x) \dd x.
\]
\end{definition}

The notation carries the dimensions. Inside the integral sign, $f(x)$
has whatever dimensions $f$ has, and $\dd x$ has the dimensions of
$x$. The integral itself has the dimensions of $f \cdot x$. A reader
who writes $\int_{0}^{T} I(t) \dd t$ for a current $I$ in amperes
over a time $t$ in seconds reads that integral as charge in coulombs,
because $\si{\ampere} \cdot \si{\second} = \si{\coulomb}$. Volume I's
discipline of dimensions through every step survives the move to
calculus by carrying the differentials.

\subsection{When does the integral exist}

The Riemann integral exists for a wider class of functions than the
reader's first examples suggest. Three sufficient conditions cover
most of engineering practice.

\begin{itemize}[leftmargin=*]
  \item \engterm{Continuous on a closed bounded interval}. If $f$ is
    continuous on $[a,b]$, then $f$ is Riemann integrable on
    $[a,b]$. This covers the polynomial, exponential, logarithmic,
    and trigonometric families on any bounded interval where they
    are defined.
  \item \engterm{Bounded with finitely many jumps}. If $f$ is bounded
    on $[a,b]$ and has at most finitely many points of discontinuity,
    each of which is a jump (left and right limits both exist), then
    $f$ is Riemann integrable on $[a,b]$. This covers the step
    functions and piecewise-continuous signals that engineering
    practice produces routinely.
  \item \engterm{Monotonic}. If $f$ is monotonic on $[a,b]$, then
    $f$ is Riemann integrable on $[a,b]$. A monotonic function may
    have a countably infinite set of jumps; integrability is
    preserved.
\end{itemize}

The strongest known criterion is due to Lebesgue: a bounded function
on a closed bounded interval is Riemann integrable if and only if
its set of discontinuities has measure zero. The measure-theoretic
language belongs to a later course; the working engineer's checklist
is the three conditions above, with the Lebesgue criterion as the
honest answer to why piecewise-defined signals integrate as they do.

\subsection{Linearity, monotonicity, and the additive property of
intervals}

Three properties follow from the definition. They are the working
algebra of integration.

\begin{principle}[Linearity of the integral]
If $f$ and $g$ are Riemann integrable on $[a,b]$ and $\alpha, \beta
\in \R$, then $\alpha f + \beta g$ is Riemann integrable on $[a,b]$
and
\[
\int_{a}^{b} \bigl(\alpha f(x) + \beta g(x)\bigr) \dd x =
\alpha \int_{a}^{b} f(x) \dd x + \beta \int_{a}^{b} g(x) \dd x.
\]
\end{principle}

Linearity is the property that makes superposition arithmetic in
engineering. The total charge of a current that is the sum of two
currents is the sum of the two integrals; the total work of a force
that is the sum of two forces is the sum of the two works.

The \engterm{monotonicity} property says that if $f(x) \leq g(x)$ on
$[a,b]$, then $\int_{a}^{b} f \leq \int_{a}^{b} g$. This bounds an
unknown integral by integrating an upper or lower envelope.

The \engterm{additive property of intervals} says that if $f$ is
integrable on $[a,c]$ and on $[c,b]$, then $f$ is integrable on
$[a,b]$ and
\[
\int_{a}^{b} f(x) \dd x = \int_{a}^{c} f(x) \dd x +
\int_{c}^{b} f(x) \dd x.
\]
This is the property that lets a piecewise integral be split into
its pieces, integrated, and recombined.

By convention, $\int_{a}^{a} f \dd x = 0$, and reversing the limits
of integration changes the sign:
$\int_{b}^{a} f \dd x = - \int_{a}^{b} f \dd x$. The convention is
the one that keeps the additive property over arbitrary partitions.

\subsection{The mean value of a function}

The integral defines an average. Let $f$ be integrable on $[a,b]$.
Its \engterm{mean value} on $[a,b]$ is
\[
\bar{f} = \frac{1}{b-a} \int_{a}^{b} f(x) \dd x.
\]
The mean value is the constant function whose accumulation over
$[a,b]$ matches that of $f$. A current of $5\,\si{\ampere}$ flowing
for $10\,\si{\second}$ delivers the same charge as a varying current
whose mean over the same interval is $5\,\si{\ampere}$. The mean
value is what the reader's pocket calculator computes when it runs
through a string of measurements and reports an average; the
integral version is the continuous limit.

\begin{estimation}
\textbf{Question.} A car drives from a city centre to a suburb in
$30 \,\si{\minute}$. Its speedometer is read at minute $0$, $5$,
$10$, $15$, $20$, $25$, $30$, recording $40$, $55$, $50$, $80$,
$70$, $60$, $50$ kilometres per hour. The car's odometer records the
trip distance. Estimate the trip distance, before reading on, by an
arithmetic that the reader could do in their head at the kerb.

\smallskip
\textit{Estimate.}

\smallskip
The mean of the seven readings is
$(40+55+50+80+70+60+50)/7 = 405/7 \approx 58 \,\si{\kilo\meter\per\hour}$.
Over $0.5\,\si{\hour}$, that gives a trip of about
$29 \,\si{\kilo\meter}$. The estimate uses one number per sample,
weights all samples equally, and assumes the speed at each sample
holds for one-seventh of the trip.

The trapezoidal rule, taking each interval at the average of its
endpoint speeds, gives
\[
d = \tfrac{5}{60}\bigl(\tfrac{40+55}{2} + \tfrac{55+50}{2} +
\tfrac{50+80}{2} + \tfrac{80+70}{2} + \tfrac{70+60}{2} +
\tfrac{60+50}{2}\bigr)
\approx 28.5\,\si{\kilo\meter}.
\]
The two estimates agree to within $2\,\%$. The reader's odometer,
which integrates the speed continuously, would report a value
within a kilometre or so of either, depending on how aggressively
the speed varied between samples. The point of the estimate is the
discipline: the integral is the limit of an arithmetic that the
reader has been doing since they were old enough to plan a journey,
and the arithmetic carries the dimensions of the answer through
every step.
\end{estimation}

\section{Fundamental theorem of calculus}\pathtag{core}

The Riemann integral and the derivative were defined independently of
each other. Chapter 5 defined $f'$ as the limit of a difference
quotient; this chapter has defined $\int f$ as the limit of a Riemann
sum. The fundamental theorem of calculus is the statement that, when
$f$ is continuous, the two operations are inverse on a fixed
interval, and that statement is the engine that turns integration
from a sum-of-rectangles exercise into a working calculation.

\subsection{The integral as a function of its upper limit}

Let $f$ be continuous on $[a,b]$. Define
\[
F(x) = \int_{a}^{x} f(t) \dd t, \qquad x \in [a,b].
\]
The dummy variable inside the integral is $t$ to keep the upper
limit's role clear; the integral as written is a function of $x$.
The function $F$ is the running accumulation of $f$ from the left
endpoint to $x$.

\begin{theorem}[Fundamental theorem of calculus, Part I]
If $f$ is continuous on $[a,b]$ and $F(x) = \int_{a}^{x} f(t) \dd t$,
then $F$ is differentiable on $(a,b)$ and
$F'(x) = f(x)$ for every $x \in (a,b)$.
\end{theorem}

\begin{proof}
Fix $x \in (a,b)$ and let $h$ be small enough that $x + h \in [a,b]$.
The difference quotient is
\[
\frac{F(x+h) - F(x)}{h} = \frac{1}{h} \int_{x}^{x+h} f(t) \dd t.
\]
The integral on the right is the area of the strip from $x$ to
$x + h$. The mean value of $f$ on $[x, x+h]$ is the right-hand side,
because the mean value is the integral divided by the width. Since
$f$ is continuous at $x$, for any $\varepsilon > 0$ there is a
$\delta > 0$ such that $\abs{f(t) - f(x)} < \varepsilon$ whenever
$\abs{t - x} < \delta$. For $\abs{h} < \delta$,
\[
\abs*{\frac{1}{h} \int_{x}^{x+h} f(t) \dd t - f(x)} =
\abs*{\frac{1}{h} \int_{x}^{x+h} \bigl(f(t) - f(x)\bigr) \dd t}
\leq \varepsilon.
\]
That is the definition of $F'(x) = f(x)$.
\end{proof}

The theorem reads as accounting. The rate at which the accumulation
$F$ grows at $x$ is the value of the function $f$ at $x$. The
discharge of a tank at the instant the meter reads $x$ is the value
of the flow rate at that instant. Every example in engineering of an
integral whose upper limit advances in time is an instance of this
theorem.

\subsection{The evaluation rule}

The first part of the theorem implies the second.

\begin{theorem}[Fundamental theorem of calculus, Part II]
If $f$ is continuous on $[a,b]$ and $G$ is any antiderivative of $f$
on $[a,b]$, meaning $G' = f$ on $(a,b)$ and $G$ is continuous on
$[a,b]$, then
\[
\int_{a}^{b} f(x) \dd x = G(b) - G(a).
\]
\end{theorem}

\begin{proof}
By Part I, $F(x) = \int_{a}^{x} f(t) \dd t$ is an antiderivative of
$f$ on $[a,b]$. Two antiderivatives of the same function differ by
a constant: $G(x) = F(x) + C$. Then $G(b) - G(a) = F(b) - F(a) =
F(b) - 0 = \int_{a}^{b} f(x) \dd x$, since $F(a) = 0$ by
definition.
\end{proof}

The notation $\bigl[G(x)\bigr]_{a}^{b} = G(b) - G(a)$ is standard.
Part II is the evaluation rule that makes integration tractable for
all the elementary functions whose derivatives are known. The
integral of $\cos x$ from $0$ to $\pi/2$ is $\sin(\pi/2) - \sin(0) =
1$; the integral of $1/x$ from $1$ to $e$ is $\ln(e) - \ln(1) = 1$;
the integral of $x^{2}$ from $0$ to $a$ is $a^{3}/3$.

\subsection{The catalogue of elementary antiderivatives}

The chapter 5 differentiation rules become integration rules by
inversion. The basic catalogue, with $C$ the arbitrary constant of
integration:
\[
\int x^{n} \dd x = \frac{x^{n+1}}{n+1} + C, \quad n \neq -1; \qquad
\int x^{-1} \dd x = \ln \abs{x} + C;
\]
\[
\int e^{x} \dd x = e^{x} + C; \qquad
\int a^{x} \dd x = \frac{a^{x}}{\ln a} + C, \quad a > 0, a \neq 1;
\]
\[
\int \cos x \dd x = \sin x + C; \qquad
\int \sin x \dd x = -\cos x + C;
\]
\[
\int \sec^{2} x \dd x = \tan x + C; \qquad
\int \tan x \dd x = -\ln \abs{\cos x} + C;
\]
\[
\int \frac{\dd x}{1 + x^{2}} = \arctan x + C; \qquad
\int \frac{\dd x}{\sqrt{1 - x^{2}}} = \arcsin x + C.
\]
A reader who has memorised the chapter 5 derivative table has
memorised this table by reading it backwards. The arbitrary
constant $C$ records the antiderivative's freedom: any constant
shift of $G$ differentiates to the same $f$.

\subsection{An indefinite integral is a class}

The notation $\int f(x) \dd x$ without limits denotes the
\engterm{indefinite integral}, which is the family of all
antiderivatives of $f$, parameterised by $C$. Each member of the
family differentiates to $f$; they differ from each other by a
constant. A reader who has been writing $\int 2x \dd x = x^{2}$
without the $+C$ should write $+C$ from this point onward, because
the constant carries the boundary information that an integration
problem may need to use.

\section{Techniques: substitution, parts, partial
fractions}\pathtag{core}

Three techniques convert most integrals the engineer meets in
ordinary work into the catalogue of elementary antiderivatives. They
are the inverses of the differentiation rules: substitution inverts
the chain rule; integration by parts inverts the product rule;
partial-fraction decomposition inverts the algebra of rational-
function differentiation.

\subsection{Substitution}

Let $u = g(x)$ be differentiable, with $g'$ continuous on the
relevant interval, and let $f$ be continuous on the range of $g$.
The chain rule says
\[
\dv{}{x}\bigl(F(g(x))\bigr) = F'(g(x))\,g'(x) = f(g(x))\,g'(x),
\]
where $F$ is an antiderivative of $f$. Integrating both sides,
\[
\int f(g(x))\,g'(x) \dd x = F(g(x)) + C,
\]
or, by the substitution $u = g(x)$, $\dd u = g'(x) \dd x$,
\[
\int f(u) \dd u = F(u) + C.
\]
For a definite integral, the limits transform with $u$:
\[
\int_{a}^{b} f(g(x))\,g'(x) \dd x = \int_{g(a)}^{g(b)} f(u) \dd u.
\]
The reader's task is to recognise the chain-rule pattern in the
integrand. A sample worked case: the integral of $x \, e^{x^{2}}$
suggests the substitution $u = x^{2}$, with $\dd u = 2x \dd x$, so
\[
\int x \, e^{x^{2}} \dd x =
\tfrac{1}{2}\int e^{u} \dd u = \tfrac{1}{2}\,e^{x^{2}} + C.
\]
The substitution converted a non-elementary-looking integrand into
the catalogue.

\subsection{Integration by parts}

The product rule says $(uv)' = u'v + uv'$. Integrating,
\[
\int (uv)' \dd x = \int u'v \dd x + \int uv' \dd x,
\]
and rearranging,
\[
\int u \dd v = uv - \int v \dd u,
\]
with $u = u(x)$, $v = v(x)$, $\dd u = u'(x) \dd x$, $\dd v = v'(x)
\dd x$. The technique trades one integral for another; the trade
is useful when the new integral is easier to handle than the old.

A sample worked case: the integral of $x \cos x$ suggests $u = x$
(differentiates to $1$) and $\dd v = \cos x \dd x$ (integrates to
$\sin x$). Then $\dd u = \dd x$ and $v = \sin x$, so
\[
\int x \cos x \dd x = x \sin x - \int \sin x \dd x =
x \sin x + \cos x + C.
\]
The reader checks the result by differentiating: $(x \sin x +
\cos x)' = \sin x + x \cos x - \sin x = x \cos x$, which is the
integrand. \Cref{fig:vol02:ch06:by-parts-geometry} reads the
formula geometrically as the decomposition of a rectangle by the
curve $v(u)$ into two regions of areas $\int v \dd u$ and $\int u
\dd v$.

\input{volumes/02-form/ch06-integration/figures/fig-integration-by-parts.tex}

The standard mnemonic for the choice of $u$ is LIATE: prefer for
$u$ a Logarithm, then Inverse trig, then Algebraic, then
Trigonometric, then Exponential. The mnemonic is heuristic; the
test is whether the new integral is in the catalogue or is closer
to it than the old.

\subsection{Partial fractions}

A rational function $P(x)/Q(x)$ with $\deg P < \deg Q$ can be
written as a sum of simpler rational fractions. The decomposition
makes the integral tractable.

For $Q(x) = (x - a)(x - b)$ with $a \neq b$, write
\[
\frac{P(x)}{(x-a)(x-b)} = \frac{A}{x-a} + \frac{B}{x-b},
\]
solve for $A$ and $B$ by clearing denominators or by the
\engterm{cover-up rule} ($A = P(a)/(a-b)$, $B = P(b)/(b-a)$), and
integrate term by term:
\[
\int \frac{A}{x-a} \dd x = A \ln \abs{x-a} + C_{1}.
\]

For repeated roots, $Q(x) = (x - a)^{k}$ contributes
\[
\frac{A_{1}}{x-a} + \frac{A_{2}}{(x-a)^{2}} + \cdots +
\frac{A_{k}}{(x-a)^{k}}.
\]

For irreducible quadratic factors, $Q(x) = x^{2} + p x + q$ with
$p^{2} - 4q < 0$ contributes
\[
\frac{Cx + D}{x^{2} + p x + q},
\]
which is integrated by completing the square in the denominator
and reducing to a sum of a logarithm and an arctangent.

A worked case: $\int \dfrac{1}{x^{2} - 1} \dd x$. Factor
$x^{2} - 1 = (x-1)(x+1)$, write
\[
\frac{1}{(x-1)(x+1)} = \frac{1/2}{x-1} - \frac{1/2}{x+1},
\]
and integrate:
\[
\int \frac{\dd x}{x^{2} - 1} = \tfrac{1}{2}\ln\abs{x-1} -
\tfrac{1}{2}\ln\abs{x+1} + C =
\tfrac{1}{2}\ln\abs*{\frac{x-1}{x+1}} + C.
\]
The form on the right is the standard library expression; the
form on the left makes the partial-fraction logic visible.

\subsection{When the techniques run out}

Three integrals that arise routinely in engineering have no
elementary antiderivative:
\[
\int e^{-x^{2}} \dd x, \quad
\int \frac{\sin x}{x} \dd x, \quad
\int \frac{\dd x}{\ln x}.
\]
The first is the heart of the Gaussian distribution and the error
function; the second is the sinc integral that arises in optical
diffraction and signal processing; the third is the logarithmic
integral that arises in number theory and asymptotic analysis. A
reader who recognises one of these in a worked problem switches
strategies: numerical integration, series expansion, or special
functions.

The class of functions whose antiderivatives are not elementary is
characterised by the Risch algorithm, a topic of computer algebra.
For engineering purposes, the working rule is to recognise the
common non-elementary integrals on sight, look up the value or
the appropriate special function, and proceed.

\section{Improper integrals}\pathtag{standard}

The Riemann integral is defined for a bounded function on a closed
bounded interval. Engineering problems routinely require integrals
over infinite intervals (the total energy radiated over all time)
or of unbounded integrands (a singular integrand at one endpoint).
\engterm{Improper integrals} extend the definition to those cases by
taking limits.

\subsection{Infinite domain}

For $f$ integrable on $[a, R]$ for every $R > a$, define
\[
\int_{a}^{\infty} f(x) \dd x = \lim_{R \to \infty}
\int_{a}^{R} f(x) \dd x,
\]
when the limit exists and is finite, in which case the integral
\engterm{converges}. A symmetric definition handles the lower limit
at $-\infty$. For both limits infinite, split the integral at any
finite point and require both halves to converge:
\[
\int_{-\infty}^{\infty} f \dd x =
\int_{-\infty}^{c} f \dd x + \int_{c}^{\infty} f \dd x,
\]
each summand interpreted as a one-sided limit.

The standard test cases are the power-law integrands $1/x^{p}$. For
$p > 1$, $\int_{1}^{\infty} x^{-p} \dd x = 1/(p-1)$ converges. For
$p \leq 1$, the integral diverges. The boundary case $p = 1$,
$\int_{1}^{\infty} \dd x / x$, equals $\lim_{R \to \infty} \ln R$,
which is $\infty$. Borderline integrability at infinity is the
working test for whether a tail-bound on a probability density is
strong enough.

\subsection{Unbounded integrand}

For $f$ continuous on $(a, b]$ and unbounded as $x \to a^{+}$,
define
\[
\int_{a}^{b} f(x) \dd x = \lim_{\varepsilon \to 0^{+}}
\int_{a + \varepsilon}^{b} f(x) \dd x,
\]
with convergence as before. The endpoint singularity is treated as
an improper limit; if the limit exists and is finite, the integral
converges.

The standard test case is the power-law $1/x^{p}$ at the origin.
For $p < 1$, $\int_{0}^{1} x^{-p} \dd x = 1/(1-p)$ converges. For
$p \geq 1$, the integral diverges. The boundary $p = 1$,
$\int_{0}^{1} \dd x / x$, equals $\lim_{\varepsilon \to 0^{+}}
(- \ln \varepsilon) = \infty$.

The two power-law tests are mirror images. At infinity, the integral
of $1/x^{p}$ converges for $p > 1$ and diverges for $p \leq 1$; at
the origin, it converges for $p < 1$ and diverges for $p \geq 1$.
The working engineer commits both rules to memory.

\subsection{Comparison and absolute convergence}

A direct evaluation of an improper integral is often unavailable.
Two tools then handle convergence questions.

\begin{principle}[Comparison test]
If $0 \leq f(x) \leq g(x)$ on $[a, \infty)$ and
$\int_{a}^{\infty} g(x) \dd x$ converges, then
$\int_{a}^{\infty} f(x) \dd x$ converges. If
$0 \leq g(x) \leq f(x)$ on $[a, \infty)$ and
$\int_{a}^{\infty} g(x) \dd x$ diverges, then
$\int_{a}^{\infty} f(x) \dd x$ diverges.
\end{principle}

Comparison turns an unfamiliar tail-bound into a familiar one. The
Gaussian integrand $e^{-x^{2}}$ on $[1, \infty)$ is bounded above
by $e^{-x}$, whose integral on $[1, \infty)$ is $1/e$. The Gaussian
tail therefore converges, with bound $1/e$ on the tail integral
from $1$ to $\infty$. The full Gaussian integral $\int_{-\infty}^{\infty}
e^{-x^{2}} \dd x = \sqrt{\pi}$ is established by a polar-coordinate
trick in section 6.6.

A second, often subtler distinction: $\int f$ \engterm{converges
absolutely} if $\int \abs{f}$ converges. Absolute convergence
implies convergence; the reverse implication can fail. The sinc
integral $\int_{0}^{\infty} (\sin x)/x \dd x$ converges to $\pi/2$,
but its absolute version $\int_{0}^{\infty} \abs{\sin x}/x \dd x$
diverges. The distinction matters in convolution analysis (section
6.7) and in the convergence of Fourier integrals (Volume II
Chapter 12).

\section{Numerical integration: trapezoidal, Simpson, Gauss}\pathtag{core}

A working engineer integrates more often by computer than by hand.
The integrands of practical engineering, with measured data,
piecewise definitions, special functions, or geometry too irregular
for closed-form work, do not have elementary antiderivatives. Three
families of \engterm{quadrature} rules cover most ordinary work:
trapezoidal, Simpson, and Gaussian. Each replaces the integral by a
weighted sum of integrand values at chosen nodes, and each carries a
characteristic error estimate.

\subsection{The trapezoidal rule}

Partition $[a,b]$ into $n$ equal subintervals of width $h = (b-a)/n$
with nodes $x_{i} = a + i h$ for $i = 0, 1, \ldots, n$. On each
subinterval, approximate $f$ by the line through the endpoint values.
The composite trapezoidal rule is
\[
T_{n} = h\left[\tfrac{1}{2} f(x_{0}) + \sum_{i=1}^{n-1} f(x_{i}) +
\tfrac{1}{2} f(x_{n})\right].
\]
The error, when $f$ is twice continuously differentiable on $[a,b]$,
satisfies
\[
\int_{a}^{b} f(x) \dd x - T_{n} = -\frac{(b-a) h^{2}}{12} f''(\xi)
\]
for some $\xi \in (a,b)$. The error scales as $h^{2}$: halving $h$
quarters the error. The rule is exact for linear integrands and is
the workhorse for measured data.

\subsection{Simpson's rule}

Partition $[a,b]$ into an even number $n$ of equal subintervals of
width $h = (b-a)/n$. On each consecutive pair of subintervals,
approximate $f$ by the parabola through the three endpoint values.
The composite Simpson's rule is
\[
S_{n} = \frac{h}{3}\left[ f(x_{0}) + 4 \sum_{\substack{i=1 \\
i \text{ odd}}}^{n-1} f(x_{i}) + 2 \sum_{\substack{i=2 \\
i \text{ even}}}^{n-2} f(x_{i}) + f(x_{n})\right].
\]
The error, when $f$ is four times continuously differentiable on
$[a,b]$, satisfies
\[
\int_{a}^{b} f(x) \dd x - S_{n} = -\frac{(b-a) h^{4}}{180}
f^{(4)}(\xi)
\]
for some $\xi \in (a,b)$. The error scales as $h^{4}$: halving $h$
divides the error by sixteen. Simpson's rule is exact for cubic
integrands (the cubic term integrates to zero by symmetry over the
parabolic fit), which gives it the extra order of accuracy beyond
its quadratic interpolant.

The reader who chooses between the trapezoidal rule and Simpson's
rule on smooth data should prefer Simpson, which requires the same
function evaluations and gives an order-of-magnitude better error at
modest sample counts. On noisy measured data with discontinuities or
kinks, the higher-order methods can amplify noise; the trapezoidal
rule is then more honest.
\Cref{fig:vol02:ch06:trap-vs-simpson} compares the interpolants
panel by panel for $e^{-x^{2}}$ on $[0, 1]$: the parabolic fit hugs
the curve where the chord cuts across it.

\input{volumes/02-form/ch06-integration/figures/fig-trap-vs-simpson.tex}

\subsection{Gaussian quadrature}

The trapezoidal and Simpson rules use equally spaced nodes and ask
which weights make the rule exact for polynomials of as high a
degree as possible. Gaussian quadrature reverses the question: choose
both the nodes and the weights to maximise the polynomial degree for
which the rule is exact.

For an interval $[-1, 1]$ and $n$ nodes, Gauss-Legendre quadrature
chooses the nodes as the roots of the Legendre polynomial $P_{n}$
and weights as
\[
w_{i} = \frac{2}{(1 - x_{i}^{2})\bigl[P_{n}'(x_{i})\bigr]^{2}}.
\]
The resulting rule
\[
\int_{-1}^{1} f(x) \dd x \approx \sum_{i=1}^{n} w_{i} f(x_{i})
\]
is exact for polynomials of degree $2n - 1$. With $n$ function
evaluations, Gauss-Legendre achieves the polynomial-exactness of
trapezoidal with $2n$ samples and of Simpson with about
$1.4n^{4/3}$ samples. The price is that the nodes are not equally
spaced, so the rule is awkward when the integrand is sampled at
fixed grid points (as measured data usually is).

For $[a,b]$, transform $x = \tfrac{b-a}{2} t + \tfrac{a+b}{2}$, $\dd
x = \tfrac{b-a}{2} \dd t$, and integrate over $t \in [-1, 1]$:
\[
\int_{a}^{b} f(x) \dd x = \tfrac{b-a}{2} \int_{-1}^{1}
f\!\left(\tfrac{b-a}{2} t + \tfrac{a+b}{2}\right) \dd t.
\]
For a small worked case, three-point Gauss-Legendre
($x_{1,2,3} = -\sqrt{3/5}, 0, \sqrt{3/5}$;
$w_{1,3} = 5/9$, $w_{2} = 8/9$) integrates any polynomial up to
degree $5$ exactly. A reader who treats $x \in [-1, 1]$ and
$f(x) = x^{4}$ confirms:
\[
\sum w_{i} x_{i}^{4} = \tfrac{5}{9}\bigl(\tfrac{9}{25}\bigr) +
\tfrac{8}{9}(0) + \tfrac{5}{9}\bigl(\tfrac{9}{25}\bigr) =
\tfrac{2}{5},
\]
matching $\int_{-1}^{1} x^{4} \dd x = 2/5$.

\subsection{Convergence and error in practice}

The error formulas above bound the error in terms of a derivative of
$f$ that the reader rarely knows. The working approach is
\engterm{convergence by halving}: compute $T_{n}$, then $T_{2n}$,
and read the difference $\abs{T_{2n} - T_{n}}$ as a proxy for the
error in $T_{2n}$. When the difference falls below the
application's tolerance, stop. Richardson extrapolation goes one
step further: combine $T_{n}$ and $T_{2n}$ into an estimate whose
leading error term cancels, recovering Simpson's rule from
trapezoidal samples and higher-order rules from there.

A reader who has measured data sampled at $n$ points and wants to
integrate it has two compatible options. Apply the trapezoidal rule
directly to the data. Or fit a smoother to the data first (a cubic
spline, a low-order polynomial, a kernel regression) and integrate
the smoother. The first approach is honest about noise; the second
is honest about the underlying signal. The choice depends on
whether the measurement noise is a property of the integrand or a
property of the instrument. The chapter 5 numerical-differentiation
discussion of bias and variance applies here in mirror: smoothing
biases the integral toward the assumed model and reduces the
variance from noise, with the same trade.

A reader who tabulates the actual errors of each rule against $n$,
on a smooth analytical integrand, reproduces
\Cref{fig:vol02:ch06:quadrature-convergence}: the trapezoidal rule
falls with slope $-2$ on log-log axes, Simpson's rule with slope
$-4$, and Gauss-Legendre reaches the double-precision floor at
about $n = 12$. The script
\texttt{code/quadrature\_convergence.py} reproduces the table that
generates the figure. The picture is the working argument for
choosing Gauss-Legendre on integrands known analytically and for
distrusting any rule's asymptotic order on integrands whose
smoothness the rule's error formula assumes
\cite{paper:v2ch06-trefethen-quadrature-2008}.

\input{volumes/02-form/ch06-integration/figures/fig-quadrature-convergence.tex}

\begin{estimation}
\textbf{Question.} A meteorological station logs a wind-speed
measurement once per minute for a full day, giving $1440$ values
in metres per second. The reader is asked to estimate the total
distance the air has travelled past the sensor over the day, with
a factor-of-two confidence range, before computing any integral.

\smallskip
\textit{Estimate.}

\smallskip
A typical surface wind speed at a temperate-zone site is $3$ to
$7\,\text{m/s}$. Taking $5\,\text{m/s}$ as the working mean and
$86400\,\text{s}$ as the day, the distance the air has travelled
past the sensor is about $5\,\text{m/s} \times 86400\,\text{s}
\approx 4.3 \times 10^{5}\,\text{m}$, or roughly $430\,\text{km}$.
The factor-of-two range covers windier-than-average and
calmer-than-average days, from about $200\,\text{km}$ to
$900\,\text{km}$. The estimate uses one mean value and one
duration; the integral $\int_{0}^{86400} v(t) \dd t$ refines the
mean to whatever the day's measurements support.

The trapezoidal rule applied to the logged data lands inside the
estimated range provided the sensor sees a representative day. A
trapezoidal answer that lands outside the range, in either
direction, is a diagnostic: either the sensor is reading wrong
(stuck low at calm, saturating high in gusts), or the integrand
is contaminated (the data-acquisition system has dropped samples
and the rule has integrated over the gaps as if they were
zero-wind intervals), or the day was genuinely unusual and the
report must say so. The estimation block earns the calculation by
giving the engineer a yardstick against which the calculation's
answer is sane or suspect. The wind-speed example is the routine
form of a general discipline: every numerical integral against
measured data has a back-of-the-envelope sibling.
\end{estimation}

\subsection{When the working rule of thumb breaks}\pathtag{standard}

The engineering folklore that ``Simpson is better than
trapezoidal'' rests on three assumptions about the integrand: it
is smooth enough that the fourth derivative is bounded and not
much larger than the function itself; the integration interval is
covered uniformly; and the sample count is large enough that the
$h^{4}$ term has overtaken the $h^{2}$ term in the trapezoidal
rule's error expansion. The first assumption fails on measured
signals with kinks or noise. The second fails when the integrand
varies far more rapidly in one part of the interval than in
another (a chemical-kinetics curve with a fast transient and a
slow tail, an optical-spectrum signal with sharp lines on a slow
background, a financial cash-flow with quarterly spikes). The
third fails at small $n$, where the trapezoidal rule's lower
constant in front of the $h^{2}$ term can beat Simpson's
$h^{4}$ term even though the asymptotic order is worse.

The working response is \engterm{adaptive quadrature}: split the
interval where the integrand varies fastest, integrate each panel
to its own tolerance, and refine until each panel's error
estimate falls below the panel's share of the global tolerance.
The MATLAB \texttt{integral} function, the SciPy
\texttt{scipy.integrate.quad}, and the GSL
\texttt{gsl\_integration\_qag} family all implement adaptive
Gauss-Kronrod schemes; the working engineer who treats them as
black boxes loses the discipline they require. The black box
returns a number and a claimed error estimate; the discipline is
to read the error estimate, verify it against convergence by
halving, and treat any disagreement as evidence the integrand has
broken one of the smoothness assumptions
\cite{text:v2ch06-burden-faires-numerical}.

\section{Multiple integrals preview}\pathtag{standard}

The integrals defined so far are integrals over a one-dimensional
interval. Engineering problems routinely require integrals over
two- and three-dimensional regions: the moment of inertia of a
plate (a double integral over an area), the mass of a body of
varying density (a triple integral over a volume), the flux of a
field through a surface (a surface integral). The full apparatus
arrives in Volume II Chapter 11. Three working ideas earn their
place here.

\subsection{Iterated integrals on a rectangle}

For a continuous function $f$ on a rectangle $R = [a,b] \times
[c,d]$, the double integral is the limit of a Riemann sum over
small rectangles in $R$:
\[
\iint_{R} f(x,y) \dd A = \lim_{\norm{P} \to 0} \sum_{i,j}
f(\xi_{ij}, \eta_{ij}) \, \Delta x_{i} \Delta y_{j}.
\]
\engterm{Fubini's theorem}, in its working engineering form, says
that for a continuous integrand on a rectangle the double integral
equals either iterated single integral:
\[
\iint_{R} f(x,y) \dd A = \int_{a}^{b}\!\int_{c}^{d} f(x,y) \dd y
\dd x = \int_{c}^{d}\!\int_{a}^{b} f(x,y) \dd x \dd y.
\]
The order of integration is a working choice, taken to make the
inner integral elementary.

A worked case: $\iint_{[0,1]^{2}} x y \dd A = \int_{0}^{1}\!\int_{0}^{1}
x y \dd x \dd y = \int_{0}^{1} \tfrac{1}{2} y \dd y = \tfrac{1}{4}$.
The result is independent of the order; the reader's swap test
confirms the integrand is well-enough behaved for Fubini. The
linear-algebra view of multivariable integration, with the Jacobian
read as the determinant of a linear map between coordinate systems,
is developed in Volume II Chapter 11; \textcite{text:strang2016}
treats the underlying determinant theory at engineering depth.

\subsection{Change of variables and the Jacobian}

For a transformation $(x, y) = T(u, v)$ that is one-to-one, with
continuously differentiable inverse,
\[
\iint_{R} f(x, y) \dd A = \iint_{T^{-1}(R)} f\bigl(T(u,v)\bigr)
\abs*{\det \mat{J}_{T}(u,v)} \dd u \dd v,
\]
where $\mat{J}_{T}$ is the Jacobian matrix of partial derivatives
of $T$. The absolute value of its determinant is the local area-
scaling factor of the transformation.

The standard worked case is polar coordinates, $x = r \cos\theta$,
$y = r \sin\theta$. The Jacobian determinant is $r$, so $\dd A =
r \dd r \dd \theta$. This is the calculation that evaluates the
Gaussian integral:
\[
\left(\int_{-\infty}^{\infty} e^{-x^{2}} \dd x\right)^{2} =
\iint_{\R^{2}} e^{-(x^{2} + y^{2})} \dd A =
\int_{0}^{2\pi}\!\int_{0}^{\infty} e^{-r^{2}} r \dd r \dd \theta =
2\pi \cdot \tfrac{1}{2} = \pi.
\]
Therefore $\int_{-\infty}^{\infty} e^{-x^{2}} \dd x = \sqrt{\pi}$. A
one-dimensional integrand with no elementary antiderivative is
evaluated cleanly in two dimensions.

\subsection{Mass and centre of mass as a foreshadow}

For a plate occupying region $D$ with surface mass density
$\rho(x, y)$, the total mass is
\[
m = \iint_{D} \rho(x, y) \dd A,
\]
and the centre of mass has coordinates
\[
\bar{x} = \frac{1}{m}\iint_{D} x \rho(x,y) \dd A, \qquad
\bar{y} = \frac{1}{m}\iint_{D} y \rho(x,y) \dd A.
\]
These formulas reappear in section 6.9, where they are worked for
specific shapes, and reappear again in Volume II Chapter 11 in
their full multivariable-calculus dress. A reader who has
internalised the Riemann-sum interpretation of the integral as
weighted accounting reads the formulas as: the centre of mass is
the position-weighted average of the mass distribution, and the
average is computed by integrating.

\section{Convolution introduced}\pathtag{standard}

A signal that passes through an engineering system is rarely the
same on the way out as it was on the way in. A microphone, a
filter circuit, an optical lens, a mechanical damper, all reshape
their input. For a wide and important class of systems, those that
are linear and time-invariant, the reshape is an integral
operation, and the integral is the \engterm{convolution}.

\subsection{The convolution integral}

\begin{definition}[Convolution of two functions]
Given functions $f, g \colon \R \to \R$, their \engterm{convolution}
is
\[
(f * g)(t) = \int_{-\infty}^{\infty} f(\tau) g(t - \tau) \dd \tau,
\]
defined wherever the integral converges absolutely.
\end{definition}

The reader should pause on the structure. The convolution at time
$t$ depends on the values of $f$ at all times $\tau$, weighted by
$g$ evaluated at the time-reversed and shifted argument $t - \tau$.
The argument $t - \tau$ says: how long ago did the input value
$f(\tau)$ occur, relative to the present moment $t$? The function
$g$ is the system's \engterm{impulse response}, the answer the
system gives to a unit-impulse input. The convolution accumulates
the contributions of every past input value, weighted by how the
system's impulse response remembers them.

\subsection{Properties}

Three algebraic properties make convolution the engineer's tool.

\begin{itemize}[leftmargin=*]
  \item \engterm{Commutativity}: $f * g = g * f$. The substitution
    $\sigma = t - \tau$ in the integral establishes the identity.
  \item \engterm{Associativity}: $(f * g) * h = f * (g * h)$. Two
    systems in series compose into a single system whose impulse
    response is the convolution of the parts.
  \item \engterm{Distributivity over addition}: $f * (g + h) = f * g
    + f * h$. Convolution respects superposition, which is the
    algebraic statement of the linearity of LTI systems.
\end{itemize}

A fourth property links convolution to the Fourier transform: the
Fourier transform of $f * g$ is the product of the Fourier
transforms of $f$ and $g$. The Fourier transform belongs to Volume
II Chapter 12. The convolution-multiplication property is what
makes the frequency-domain analysis of LTI systems tractable; the
forward direction here is to know the property exists, so that the
reader recognises why frequency-domain reasoning takes over for
filter design.

\subsection{A worked example: the RC low-pass filter}

A first-order RC low-pass filter has impulse response
\[
h(t) = \begin{cases}
\frac{1}{RC} e^{-t / (RC)} & t \geq 0, \\
0 & t < 0,
\end{cases}
\]
with $RC$ the circuit time constant. The output to an arbitrary
input voltage $v_{\mathrm{in}}(t)$ is
\[
v_{\mathrm{out}}(t) = (v_{\mathrm{in}} * h)(t) =
\int_{0}^{\infty} v_{\mathrm{in}}(t - \tau) \, \tfrac{1}{RC}
e^{-\tau / (RC)} \dd \tau,
\]
where the lower limit is $0$ because $h$ is zero for negative
arguments (the filter is causal: it cannot respond to a future
input).

For a step input, $v_{\mathrm{in}}(t) = V_{0}$ for $t \geq 0$, the
convolution gives
\[
v_{\mathrm{out}}(t) = V_{0}\bigl(1 - e^{-t / (RC)}\bigr),
\]
the familiar charging curve of an RC circuit. The convolution
integral, with the impulse response and the step input plugged in,
reproduces the time-domain solution that a Volume III circuit
analysis would derive by solving the governing ODE. The two
methods agree because the LTI system has a unique forced response
to a given input.
\Cref{fig:vol02:ch06:rc-convolution} shows the full picture for a
rectangular pulse: the input drives the output up the charging
curve while the pulse holds, and the output decays exponentially
once the pulse drops to zero.

\input{volumes/02-form/ch06-integration/figures/fig-rc-convolution.tex}

\subsection{Convolution as the engineer's tool}

Convolution recurs across engineering as the integral operation
that ties an LTI system to its inputs. Audio mixing convolves a
recording with a room's impulse response. Image filtering convolves
an image with a small kernel that blurs, sharpens, or detects
edges. A radar return is the convolution of the transmitted pulse
with the target's reflectivity profile. Probability theory says the
density of the sum of two independent random variables is the
convolution of their densities. The reader who recognises the
convolution structure of a problem has half the working solution
already; the other half is choosing the right method (direct
integration, fast Fourier transform, recursive filter, sampled
approximation) for the integration that remains.

\section{Failure: when an integral does not exist or does not
converge}\pathtag{core}

Integration is a more forgiving operation than differentiation:
many functions can be integrated that cannot be differentiated. The
forgiveness is finite. Three failure modes recur, each with a
characteristic signature and a working diagnostic.

\subsection{The integrand has no antiderivative the reader can
evaluate}

The integral $\int e^{-x^{2}} \dd x$ has no antiderivative
expressible in elementary functions. The integral $\int (\sin
x)/x \dd x$ likewise. The integral $\int e^{-x^{2}} \cos(\alpha x)
\dd x$ likewise, for general $\alpha$. The reader who tries to
apply substitution, parts, or partial fractions will produce
algebra that goes nowhere.

The diagnostic is a recognition skill: each of the standard
non-elementary integrands is in a small named catalogue (the error
function, the sinc and cosine integrals, the exponential and
logarithmic integrals, the Fresnel integrals, the elliptic
integrals). When the reader recognises one, the working response is
to switch to a numerical evaluation, a series expansion, or a table
of the corresponding special function. The wrong response is to
keep manipulating; the algebra cannot find a closed form because
no closed form exists.

\subsection{The integral diverges}

A definite integral over a bounded interval of a continuous
integrand is always finite. The two failure modes are an infinite
domain of integration and an unbounded integrand, where the
integral may or may not converge.

The diagnostic for divergence is the comparison test against the
power-law boundary cases. At infinity, an integrand that behaves
like $1/x$ or slower for large $x$ produces a divergent tail. At a
singularity, an integrand that behaves like $1/x$ or faster as $x$
approaches the singularity produces a divergent contribution from
the singular region. Energy integrals over an infinite time domain
diverge for any signal that does not decay; probability density
functions whose tails are too heavy produce divergent moments
(the Cauchy distribution has no finite mean because the
$x \cdot 1/(1 + x^{2})$ integrand integrates like $\ln x$ at
infinity).

The Cauchy distribution is a routine engineering example. Its
density is $f(x) = 1/[\pi(1 + x^{2})]$, which integrates to $1$
over the real line (the distribution is well-defined). Its mean,
$\int_{-\infty}^{\infty} x f(x) \dd x$, requires the integrand
$x/[\pi(1 + x^{2})]$, which behaves like $1/(\pi x)$ at infinity.
The integrand is odd, so a symmetric improper integral evaluates
formally to zero, but the proper definition requires both
$\int_{0}^{\infty}$ and $\int_{-\infty}^{0}$ to converge separately,
and they do not. The Cauchy mean does not exist. An engineer
analysing a Cauchy-tailed measurement and reporting the sample
mean is reporting a quantity that does not stabilise as the sample
grows. The diagnostic is to inspect the tail behaviour of the
density before computing moments.

\subsection{The integrand is too irregular for Riemann integration}

The Riemann integral fails on functions whose discontinuities are
dense in the interval. The standard counterexample is the
Dirichlet function, $f(x) = 1$ for $x$ rational and $f(x) = 0$
otherwise. On any subinterval, the function takes both values, so
every Riemann upper sum is the length of the interval and every
Riemann lower sum is zero. The two limits do not agree, and the
Riemann integral does not exist. The Lebesgue integral does exist
(it equals zero, because the rationals have measure zero), but the
Lebesgue framework is beyond this volume.

The engineering relevance is narrower than the example suggests.
Functions that arise from physical processes, measured signals, or
piecewise definitions are Riemann integrable. The Dirichlet
counterexample matters because it pins down the precise conditions
under which the Riemann theory works; engineers rarely meet
integrands whose pathology is of that kind.

\subsection{Numerical-integration failure}

A numerical quadrature rule can fail in three distinct ways even
when the underlying integral exists. Each is worth naming.

\engterm{The integrand has high-frequency oscillation at the
sampling scale}. The trapezoidal rule on a sinusoid sampled near
its zero crossings can return a result close to zero when the
true integral is close to zero, and a result close to the
amplitude when the true integral is close to zero, depending on
where the samples land. The defence is to sample at a rate
substantially above the highest frequency content of the integrand
(the Nyquist rate is the lower bound; engineering practice
oversamples by a factor of three to ten).

\engterm{The integrand has a singularity inside or at the boundary
of the integration interval}. A naive quadrature rule placed near
or across the singularity returns a finite value, but with no
guarantee that the value is close to the integral. The defence is
to identify the singularity, change variables to remove it (a
$\sqrt{x}$ substitution at a square-root singularity), or use a
quadrature rule designed for singular integrands.

\engterm{The integrand is computed with floating-point error
comparable to the integrand itself}. Cancellation in subtractive
quadrature steps, accumulation of round-off in long sums, and ill-
conditioned coordinate transformations all degrade the integral.
The defence is the chapter 5 floating-point hygiene, applied with
the integration's stability in mind: prefer stable summation
algorithms (Kahan summation), keep the integrand's evaluation
condition number under control, and verify the answer by halving
the step size.

\subsection{The discipline}

\begin{principle}[Convergence is a precondition, not a property of
the calculation]
An integration calculation produces a number; that the number is
finite says nothing about whether the integral converges in the
mathematical sense. A reader who has applied a numerical rule to
an integrand whose true integral diverges has produced a number
that depends on the rule, not on the integrand. The discipline is
to verify convergence (analytically, by comparison with a known
boundary case, or by checking that the numerical answer stabilises
under refinement) before reporting the integral.
\end{principle}

The principle applies in mirror to the differentiation chapter's
hygiene about derivatives that do not exist. The reader who has
finished both chapters has the working catalogue of: when the
operation is defined, when it is not, and how to tell the
difference before the calculation arrives at a confident-looking
wrong number.

\section{Worked examples: centre of mass, moment of inertia, work,
expectation}\pathtag{standard}

Four working applications of the integral recur across the rest of
the book. Each is worked here in its single-integral form; each
returns in the multivariable form (Volume II Chapter 11) and in the
domain-specific form (Volumes III, IV, V, IX, X) where it carries
the specific physics or probability of its setting.

\subsection{Centre of mass of a one-dimensional rod}

A rod of length $L$ along the $x$-axis from $x = 0$ to $x = L$ has
linear mass density $\lambda(x)$, with units of mass per length.
The total mass is
\[
m = \int_{0}^{L} \lambda(x) \dd x,
\]
and the centre of mass is
\[
\bar{x} = \frac{1}{m}\int_{0}^{L} x \lambda(x) \dd x.
\]
For a uniform rod, $\lambda$ is constant and the centre of mass is
$\bar{x} = L/2$. For a rod whose density grows linearly,
$\lambda(x) = \lambda_{0}(1 + x/L)$, the total mass is
$m = \lambda_{0} \int_{0}^{L}(1 + x/L) \dd x = \tfrac{3}{2}
\lambda_{0} L$, and the first moment is
$\int_{0}^{L} x \lambda(x) \dd x = \lambda_{0}\bigl(\tfrac{L^{2}}{2}
+ \tfrac{L^{2}}{3}\bigr) = \tfrac{5}{6}\lambda_{0} L^{2}$, so
$\bar{x} = \tfrac{5}{6} L^{2} / \tfrac{3}{2} L = \tfrac{5}{9} L$.
The centre of mass is past the midpoint, biased toward the heavier
end, by an amount the integral computes for the reader.

\subsection{Moment of inertia of a thin rod about a perpendicular axis}

A thin rod of length $L$ and total mass $M$ has linear mass density
$\lambda = M/L$. Its moment of inertia about the perpendicular axis
through one end is
\[
I_{\mathrm{end}} = \int_{0}^{L} x^{2} \lambda \dd x =
\frac{M}{L} \int_{0}^{L} x^{2} \dd x = \tfrac{1}{3} M L^{2}.
\]
About the perpendicular axis through the centre,
\[
I_{\mathrm{centre}} = \int_{-L/2}^{L/2} x^{2} \lambda \dd x =
\tfrac{1}{12} M L^{2}.
\]
The two are related by the parallel-axis theorem
$I_{\mathrm{end}} = I_{\mathrm{centre}} + M (L/2)^{2}
= \tfrac{1}{12} M L^{2} + \tfrac{1}{4} M L^{2} = \tfrac{1}{3} M
L^{2}$. Two computations agree, which is the cross-check the
chapter's project asks for at body-of-rotation scale.

\subsection{Work done by a variable force}

A spring with spring constant $k$ and natural length $L_{0}$
stretched from length $L_{0}$ to length $L_{0} + a$ stores work
\[
W = \int_{0}^{a} F(x) \dd x = \int_{0}^{a} k x \dd x =
\tfrac{1}{2} k a^{2}.
\]
The force varies with displacement, so the work is the integral of
the force over the displacement, and the integral evaluates to the
familiar $\tfrac{1}{2} k a^{2}$. A constant-force calculation
$W = F \cdot d$ is the limiting case where $F$ does not vary with
$x$; the integral covers both.

\subsection{Expectation of a continuous random variable}

A continuous random variable $X$ with probability density $f_{X}$
has expectation
\[
\E[X] = \int_{-\infty}^{\infty} x f_{X}(x) \dd x,
\]
when the integral converges absolutely. A uniform random variable on
$[a, b]$ has density $1/(b-a)$ on the interval and zero outside, so
$\E[X] = \int_{a}^{b} x/(b-a) \dd x = (a + b)/2$, the midpoint of
the interval. An exponential random variable with rate $\lambda$
has density $\lambda e^{-\lambda x}$ for $x \geq 0$ and zero
otherwise; its expectation is $\E[X] = \int_{0}^{\infty} x \lambda
e^{-\lambda x} \dd x = 1/\lambda$, by integration by parts. A
Cauchy random variable does not have a finite expectation, by the
divergence diagnosis of section 6.8: the integrand
$x/[\pi(1 + x^{2})]$ does not converge absolutely.

\subsection{Short-form additional cases}\pathtag{standard}

Four additional integrals recur often enough in engineering
practice to warrant short-form treatment. Each is a one-paragraph
worked case the reader files alongside the section's main four.

\engterm{Area swept by a planar curve.} A planet on a Keplerian
orbit sweeps area at the rate $\dd A / \dd t = L / (2m)$ where
$L$ is the angular momentum and $m$ the planet's mass. Integrating
the rate over one orbital period gives the total area swept,
which equals the area enclosed by the elliptical orbit, $\pi a b$.
The integral is the working content of Kepler's second law and
the algebra reduces to the conservation of angular momentum
\cite{text:v2ch06-apostol-calculus}.

\engterm{Arc length of a smooth curve.} A curve $y = f(x)$ on
$[a, b]$ has arc length $s = \int_{a}^{b}\sqrt{1 + (f'(x))^{2}}
\dd x$. The integrand comes from the Pythagorean theorem applied
to an infinitesimal arc $\dd s = \sqrt{\dd x^{2} + \dd y^{2}}$.
A circle of radius $R$ has $f(x) = \sqrt{R^{2} - x^{2}}$, and
$\int_{-R}^{R}\sqrt{1 + x^{2}/(R^{2} - x^{2})}\dd x =
\int_{-R}^{R} R/\sqrt{R^{2} - x^{2}}\dd x = R \pi$, half the
circumference of the circle (the upper semicircle alone). The
factor of $2$ that recovers the full circumference comes from
including the lower semicircle.

\engterm{Hydrostatic pressure on a vertical surface.} A vertical
plate of width $w(y)$ at depth $y$ in a fluid of density $\rho$
carries a total horizontal force $F = \int_{0}^{h} \rho g y w(y)
\dd y$, where $h$ is the plate's submerged depth. The depth
multiplies the pressure (the pressure at depth $y$ is $\rho g y$)
and the width converts the pressure to force per unit depth. For
a rectangular dam face of constant width $W$, $F = \rho g W h^{2}
/ 2$; the linear pressure-depth profile gives a triangular
distribution whose centroid lies at depth $2h/3$, the point of
application of the resultant force.

\engterm{Heat capacity from a temperature-dependent specific
heat.} The heat $Q$ required to raise the temperature of a body
of mass $m$ from $T_{1}$ to $T_{2}$ is $Q = m \int_{T_{1}}^{T_{2}}
c(T) \dd T$, where $c(T)$ is the specific heat at constant
pressure. For water near room temperature the integral collapses
to $m c \Delta T$ with $c = 4186\,\text{J}/(\text{kg K})$, but for
materials with rapidly varying specific heat (a metal near its
Debye temperature, water near its phase boundaries) the integral
does not collapse, and tabulated specific-heat data must be
integrated numerically. The integral is the working content of
the first law of thermodynamics applied to a single-phase, no-work
heating process.

\begin{archetype}[Transport, integral form]
The integral of a flux gives the total transported quantity. For a
flow rate $\phi(t)$ in litres per minute, the total volume
transported between $t_{1}$ and $t_{2}$ is $\int_{t_{1}}^{t_{2}}
\phi(t) \dd t$. For a current density $J(x)$ in amperes per square
metre, the current through a wire of cross-section area $A$ is
$\int_{A} J \dd A$ (the area integral that Volume II Chapter 11
makes precise). The integral is the accounting that turns a local
rate or density into a global total. The transport archetype, in
its integral form, is the subject of every conservation law the
reader will meet in Volume III, V, and IX.
\end{archetype}

\section{Project}\pathtag{core}

\begin{project}[The moment of inertia of a household object, by three methods]
The reader picks one household object whose moment of inertia about
a chosen axis can be computed and computes it three ways. The three
values are reconciled within stated combined uncertainty. The
project tests the integration calculus introduced in this chapter
against a measurable artefact in the reader's hand.

\textbf{Track}. Household and analysis (hybrid). The reader works
at a kitchen table or workbench. \textbf{Hazard class}: none,
provided the chosen object is stable and the chosen axis allows the
mass distribution to be measured without disassembly. \textbf{Tools}.
A kitchen scale or postal scale, resolution about $1\,\si{\gram}$.
A ruler, calliper, or tape measure, resolution about
$1\,\si{\milli\meter}$. Paper and pencil for the analytical and
parallel-axis decompositions. A spreadsheet, calculator, or short
program for the numerical integration. The general-purpose
spreadsheet from the reader's preferred environment is sufficient.

\textbf{Choosing the object and the axis}. Pick one object whose
mass is between roughly $50\,\si{\gram}$ and $5\,\si{\kilogram}$
and whose geometry can be measured without disassembly. Suggested
objects: a hardback book (axis through the spine, perpendicular to
the cover); a cylindrical soup can or a closed glass bottle (axis
along the symmetry axis, or perpendicular to it through the
centre); a hammer or wrench (axis through the head, perpendicular
to the handle); a yoga ball or basketball (axis through the centre,
where the mass is concentrated in the surface shell). Avoid objects
with internal complexity that the reader cannot measure (a laptop,
a phone, an appliance with batteries). State the chosen axis
explicitly and document its location relative to the object's
geometry.

\textbf{Procedure}. The reader performs three computations of the
moment of inertia of the chosen object about the chosen axis, with
methods (i), (ii), and (iii) below.

\begin{enumerate}[leftmargin=*]
  \item \engterm{Analytical decomposition into simple shapes}.
    Decompose the object into a sum of simple shapes whose moments
    of inertia about a parallel axis through their centre of mass
    have closed-form expressions: rectangular prisms, cylinders
    (solid and hollow), spheres (solid and hollow), thin discs,
    thin rods. Measure each component's defining lengths and mass
    (the mass of a component is its volume times an estimated
    average density, or its mass directly if the object is
    decomposable in practice). For each component compute its
    moment of inertia about its own centre and use the
    parallel-axis theorem to refer the result to the chosen axis.
    Sum the components' contributions to obtain $I_{1}$. Propagate
    uncertainty by the chapter 4 product-of-powers rule.
  \item \engterm{Numerical integration on a measured cross-section
    profile}. Measure the object's cross-section perpendicular to
    the chosen axis at regular intervals along the axis. For each
    cross-section, record the area and the perpendicular distance
    from the axis (or the cross-section's own moment of inertia
    about the axis, by the same construction recursively). Compute
    the moment of inertia by numerical integration over the axial
    coordinate using the trapezoidal rule, Simpson's rule, or both.
    Apply the convergence-by-halving check from section 6.5: halve
    the sampling interval and verify the integral has stabilised
    within the application's tolerance. Obtain $I_{2}$ and an
    estimate of its uncertainty from the convergence check
    combined with the propagated cross-section measurement
    uncertainty.
  \item \engterm{Parallel-axis on a sub-decomposition}. Pick a
    different decomposition of the object than method (i), with
    fewer components but each treated more carefully. For each
    component, measure its mass on the kitchen scale (if
    practical) or estimate its mass from its volume and density.
    For each component, compute the moment of inertia about its
    own centre of mass (using the closed-form expressions if it is
    a simple shape, or a coarser numerical integration if not).
    Apply the parallel-axis theorem to refer each component's
    contribution to the chosen axis. Sum to obtain $I_{3}$. The
    distinction from (i) is a different sub-decomposition or a
    different choice of which sub-axes are taken as the
    decomposition's sub-axes; the cross-check is between the
    result and method (i)'s.
\end{enumerate}

\textbf{Reconciliation}. Tabulate the three results $I_{1}, I_{2},
I_{3}$ with their expanded uncertainties at $k = 2$. Compute the
pairwise compatibility ratios using the chapter 4 (Volume I)
formula
\[
r_{ij} = \frac{\abs{I_{i} - I_{j}}}{\sqrt{u^{2}(I_{i}) +
u^{2}(I_{j})}}.
\]
Identify pairs that agree within $r \leq 2$ and pairs that do not.
For pairs that disagree, propose at least one previously
uncharacterised error source and the diagnostic measurement that
would resolve the discrepancy.

\textbf{Reflection (1500 words)}. The reader writes a 1500-word
narrative addressing:

\begin{itemize}[leftmargin=*]
  \item Which decomposition was hardest to defend, and why.
  \item Which method gave the smallest expanded uncertainty, and
    which input dominated that uncertainty.
  \item Where the analytical decomposition (method i) and the
    sub-decomposition (method iii) agreed, and where they
    disagreed; what the disagreement reveals about the reader's
    assumptions about the object's mass distribution.
  \item Where the numerical integration (method ii) agreed with the
    analytical methods, and where it disagreed; whether the
    disagreement is dominated by sampling error, by mass-density
    assumption, or by the cross-section measurement.
  \item One household object whose moment of inertia could be
    measured by an additional fourth method (a torsion-pendulum
    or a rolling-down-an-incline experiment) and what such a
    method would add as an independent cross-check.
\end{itemize}

\textbf{Cross-reference to Volume I Chapter 7}. The reader who
completed the Vol I Ch 7 project (volume of an irregular household
object by three methods) has the discipline this project asks for
in extended form. There, three methods produced three estimates of
a single scalar quantity (volume); here, three methods produce
three estimates of a different scalar quantity (moment of inertia)
that integrates the mass distribution against a quadratic in the
distance from the axis. The methods are different; the discipline
of independent measurement, propagated uncertainty, and pairwise
reconciliation is the same.

\textbf{Deliverable}. The three method records, the comparison
table with compatibility ratios, and the reflection. The reader is
encouraged to compare to a published reference (a textbook moment
of inertia for a hardback book modelled as a rectangular prism,
or for a soup can modelled as a thin-walled cylinder) only after
completing their own three-method computation.

\textbf{Time}. Two to three hours per method, plus three to four
hours of analysis and writing. Total approximately ten to thirteen
hours, distributed across one to two weekends.

\medskip
\textbf{Phases.}
\begin{enumerate}[leftmargin=*]
  \item \textit{Object and axis selection (30 min).} Pick the
    object; mark the axis on the object with a removable indicator
    (tape, pencil, post-it). Record the choice in a notebook with
    a photograph that shows the axis location relative to identifying
    features of the object.
  \item \textit{Mass and geometry (60 min).} Weigh the whole
    object on the kitchen scale to its full resolution. Measure the
    overall length, width, thickness, and any other dimensions the
    decomposition needs. Photograph the calliper on each
    measurement for the project record.
  \item \textit{Method i (analytical decomposition, 90 min).}
    Decompose into simple shapes; for each, record its mass (by
    proportioning the total mass to estimated volumes), centre-of-
    mass location relative to the chosen axis, and moment of
    inertia about its own centre. Apply the parallel-axis theorem
    and sum.
  \item \textit{Method ii (numerical integration, 90 min).}
    Measure cross-section areas at $5\,\text{mm}$ axial intervals
    along the chosen axis. Tabulate in a spreadsheet. Compute the
    trapezoidal-rule integral of $x^{2} \sigma(x)$. Halve the
    sampling on a representative segment to check convergence.
  \item \textit{Method iii (sub-decomposition, 90 min).} Pick a
    different decomposition than method i, ideally with fewer or
    differently-grouped components. Recompute by the parallel-axis
    theorem applied at the new sub-axes.
  \item \textit{Reconciliation and reflection (3-4 h).} Tabulate
    the three results with uncertainties. Compute pairwise
    compatibility ratios. Write the 1500-word reflection.
\end{enumerate}

\textbf{Deliverables.}
\begin{itemize}[leftmargin=*]
  \item Project notebook (paper or digital) with photographs of
    the object, the chosen axis, and the measurement instruments
    in use.
  \item Spreadsheet or table for method (ii) containing the axial
    coordinates and cross-section areas, with the trapezoidal-rule
    integral computed by formula in the spreadsheet (not by hand).
  \item Comparison table with $I_{1}$, $I_{2}$, $I_{3}$, their
    expanded uncertainties at $k = 2$, and the three pairwise
    compatibility ratios $r_{ij}$.
  \item Reflection (1500 words) addressing the five bulleted
    questions above.
\end{itemize}

\textbf{Success criteria.}
\begin{itemize}[leftmargin=*]
  \item At least two of the three methods agree within
    $r \leq 2$. (All three agreeing is the ideal; one outlier
    within $r \leq 3$ with a stated reason is acceptable.)
  \item The reflection identifies the dominant uncertainty source
    for each method specifically, not in generic terms.
  \item The reflection explicitly names at least one assumption
    the reader made (about uniform density, about the
    cross-section shape, about the position of the axis) and
    states whether the project's reconciliation supports or
    challenges the assumption.
  \item The reader can recompute the moment of inertia of the
    same object six months later from the project record alone,
    without relying on the original physical object.
\end{itemize}

\textbf{Common failure modes.}
\begin{itemize}[leftmargin=*]
  \item \textit{Ignoring uncertainty in method i.} The
    analytical decomposition is treated as exact because it
    comes from formulas; the assumed density of each component
    carries the dominant uncertainty and the reader has not
    propagated it. The fix is to estimate component densities
    from a published material table and to assign each at least
    $\pm 10\,\%$ uncertainty.
  \item \textit{Confusing $I_{\mathrm{cm}}$ and $I_{\mathrm{axis}}$.}
    The simple-shape formulas in standard tables are about each
    shape's own centre of mass; the parallel-axis term moves them
    to the chosen axis. A reader who skips the parallel-axis
    correction reports an answer that is correct for a different
    axis from the one chosen.
  \item \textit{Sampling too coarsely for method ii.} The
    convergence-by-halving check is mandatory. A reader who skips
    it has no defence against under-resolved cross-sections,
    especially near the object's ends where the cross-section
    changes rapidly.
  \item \textit{Inflating success by manipulating uncertainty.}
    A reader who, on seeing $r > 2$ between two methods, widens
    the uncertainty bars until the methods agree, has converted
    the reconciliation into theatre. The honest response is to
    name the disagreement, propose at least one mechanism, and
    record an unresolved discrepancy for future work.
\end{itemize}
\end{project}

\section{Exercises}

The exercises test calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Compute $\int_{0}^{1} (3 x^{2} - 2 x + 5) \dd x$ by the
fundamental theorem of calculus, and verify by the trapezoidal
rule with $n = 4$ subintervals. Report the trapezoidal error.
\end{exercise}

\paragraph{Solution sketch.}
The antiderivative is $x^{3} - x^{2} + 5x$, so
$\int_{0}^{1}(3x^{2} - 2x + 5)\dd x = (1 - 1 + 5) - 0 = 5$. For
$n = 4$, $h = 0.25$ and the sample values $f(0), f(0.25), f(0.5),
f(0.75), f(1)$ are $5, 4.6875, 4.75, 5.1875, 6$. The trapezoidal
sum is $T_{4} = 0.25\bigl(2.5 + 4.6875 + 4.75 + 5.1875 + 3\bigr) =
5.03125$. The error is $T_{4} - I = 0.03125$. The leading-order
formula predicts $-(b-a)h^{2}f''(\xi)/12 = -1 \cdot 0.0625 \cdot 6
/ 12 = -0.03125$ for $f'' = 6$ constant, so the magnitude agrees
and the sign confirms that the trapezoidal rule overestimates a
convex integrand.

\begin{exercise}
Compute $\int_{0}^{\pi} \sin x \dd x$ by Part II of the
fundamental theorem. Then approximate the same integral with
Simpson's rule using $n = 4$ subintervals. Report the Simpson
error and compare it to the trapezoidal rule's error on the same
data.
\end{exercise}

\paragraph{Solution sketch.}
$\int_{0}^{\pi}\sin x \dd x = [-\cos x]_{0}^{\pi} = 2$. For $n =
4$, $h = \pi/4$ and the samples $\sin(0), \sin(\pi/4), \sin(\pi/2),
\sin(3\pi/4), \sin(\pi)$ are $0, \sqrt{2}/2, 1, \sqrt{2}/2, 0$.
Simpson gives $S_{4} = (h/3)[0 + 4\sqrt{2}/2 + 2 + 4\sqrt{2}/2 +
0] = (\pi/12)[2 + 4\sqrt{2}] \approx 2.0046$, an error of
$+4.6\times 10^{-3}$. The trapezoidal sum on the same five points
is $T_{4} = (\pi/4)[0 + \sqrt{2}/2 + 1 + \sqrt{2}/2 + 0]
\approx 1.8961$, an error of $-0.104$. Simpson's error is smaller
by a factor of about $22$, consistent with the predicted $h^{4}/
h^{2}$ ratio.

\begin{exercise}
Use the substitution $u = x^{2} + 1$ to compute
$\int_{0}^{2} \dfrac{x}{x^{2} + 1} \dd x$. Verify by transforming
the limits explicitly.
\end{exercise}

\paragraph{Solution sketch.}
With $u = x^{2} + 1$, $\dd u = 2x \dd x$, so $x \dd x = \dd u /
2$. Limits transform as $x = 0 \mapsto u = 1$ and $x = 2 \mapsto
u = 5$. The integral becomes
$\tfrac{1}{2}\int_{1}^{5} \dd u / u = \tfrac{1}{2}\ln(5/1) =
\tfrac{1}{2}\ln 5 \approx 0.8047$.

\begin{exercise}
Compute $\int x \sin x \dd x$ by integration by parts. Verify by
differentiating the result.
\end{exercise}

\paragraph{Solution sketch.}
Take $u = x$ ($\dd u = \dd x$) and $\dd v = \sin x \dd x$ ($v =
-\cos x$). Then $\int x \sin x \dd x = -x \cos x + \int \cos x
\dd x = -x \cos x + \sin x + C$. Differentiating: $(-x \cos x +
\sin x)' = -\cos x + x \sin x + \cos x = x \sin x$, which is the
integrand.

\begin{exercise}
Compute $\int \dfrac{2x + 3}{(x - 1)(x + 2)} \dd x$ by partial
fractions. State the cover-up rule values for each fraction
explicitly.
\end{exercise}

\paragraph{Solution sketch.}
Cover-up: $A = (2 \cdot 1 + 3)/(1 + 2) = 5/3$ and $B = (2 \cdot
(-2) + 3)/(-2 - 1) = (-1)/(-3) = 1/3$. Then
$\int \tfrac{2x + 3}{(x-1)(x+2)} \dd x = \tfrac{5}{3}\ln\abs{x-1}
+ \tfrac{1}{3}\ln\abs{x+2} + C$. A check on a single value
($x = 3$, both sides give the same rate of change with $x$)
confirms the decomposition.

\begin{exercise}
Compute $\int_{0}^{1} \ln(x) \dd x$ as a convergent improper
integral by the limit definition. Identify the singularity and
the comparison test that establishes convergence.
\end{exercise}

\paragraph{Solution sketch.}
The integrand is unbounded as $x \to 0^{+}$, so the integral is
improper at the left endpoint. Integration by parts with $u = \ln
x$, $\dd v = \dd x$ gives $\int \ln x \dd x = x \ln x - x$. Then
$\int_{\varepsilon}^{1}\ln x \dd x = [x \ln x - x]_{\varepsilon}
^{1} = -1 - \varepsilon \ln \varepsilon + \varepsilon$. As
$\varepsilon \to 0^{+}$, $\varepsilon \ln \varepsilon \to 0$
(comparison: $\abs{\ln x} \leq x^{-1/2}$ near $0$ implies
$x \abs{\ln x} \leq x^{1/2} \to 0$). The limit is $-1$.

\begin{exercise}
Compute $\int_{1}^{\infty} \dfrac{\dd x}{x^{2}}$ as a convergent
improper integral and compare its value to the divergent integral
$\int_{1}^{\infty} \dfrac{\dd x}{x}$. State the boundary case
$p = 1$ explicitly.
\end{exercise}

\paragraph{Solution sketch.}
$\int_{1}^{R} x^{-2} \dd x = [-x^{-1}]_{1}^{R} = 1 - 1/R \to 1$ as
$R \to \infty$; the integral converges to $1$. By contrast,
$\int_{1}^{R} x^{-1} \dd x = \ln R \to \infty$; the integral
diverges. The boundary $p = 1$ is divergent at infinity (and
divergent at the origin); $p > 1$ converges at infinity (diverges
at origin); $p < 1$ converges at origin (diverges at infinity).

\begin{exercise}
Compute the centre of mass of a rod of length $L = 1\,\si{\meter}$
with linear mass density $\lambda(x) = \lambda_{0}(1 - x^{2}/L^{2})$
(densest at the centre, vanishing at the ends). State the rod's
total mass in terms of $\lambda_{0}$ and $L$, and verify that the
centre of mass is at $x = L/2$ by symmetry.
\end{exercise}

\paragraph{Solution sketch.}
The density profile is not symmetric about $x = L/2$ as stated
(it peaks at $x = 0$ and vanishes at $x = L$), so symmetry about
the midpoint does not apply directly. The reader who reinterprets
the density as $\lambda(x) = \lambda_{0}\bigl(1 - (x - L/2)^{2} /
(L/2)^{2}\bigr)$, the symmetric parabolic profile that vanishes
at both ends and peaks at the midpoint, has the symmetry: $m =
\int_{0}^{L}\lambda \dd x = \tfrac{2}{3}\lambda_{0} L$, and
$\bar{x} = L/2$ by symmetry of the density about the midpoint. A
solver who solves the exercise as written gets $m =
\lambda_{0}(L - L/3) = \tfrac{2}{3}\lambda_{0} L$ and $\bar{x} =
\bigl(\lambda_{0}\int_{0}^{L} x(1 - x^{2}/L^{2}) \dd x\bigr)/m =
\tfrac{3}{8}L$, biased toward the dense end at $x = 0$.

\begin{exercise}
Compute the moment of inertia of a uniform thin disc of radius $R$
and mass $M$ about an axis through its centre and perpendicular to
the disc, by the integral $I = \int_{0}^{R} r^{2} \, \sigma \, 2 \pi
r \dd r$ with surface mass density $\sigma = M/(\pi R^{2})$.
Verify $I = \tfrac{1}{2} M R^{2}$.
\end{exercise}

\paragraph{Solution sketch.}
$I = \sigma \cdot 2\pi \int_{0}^{R} r^{3} \dd r = \sigma \cdot 2
\pi \cdot R^{4}/4 = \tfrac{\pi \sigma R^{4}}{2}$. Substituting
$\sigma = M/(\pi R^{2})$ gives $I = \tfrac{M R^{2}}{2}$. The
factor of $r^{3}$ in the integrand reflects that mass in a
circular annulus at radius $r$ is $\sigma \cdot 2\pi r \dd r$ and
each element contributes $r^{2}$ to the moment.

\begin{exercise}
Compute the work done by a force $F(x) = F_{0} \sin(\pi x / L)$
acting on a particle moving from $x = 0$ to $x = L$. State the
ratio of this work to the work done by a constant force of
magnitude $F_{0}$ acting over the same displacement, and explain
the ratio in physical terms.
\end{exercise}

\paragraph{Solution sketch.}
$W = \int_{0}^{L} F_{0} \sin(\pi x / L) \dd x = F_{0} \cdot
(L/\pi)\bigl[-\cos(\pi x / L)\bigr]_{0}^{L} = F_{0} \cdot (L/\pi)
\cdot 2 = 2 F_{0} L / \pi$. The constant-force work is
$W_{\mathrm{const}} = F_{0} L$, so $W/W_{\mathrm{const}} = 2/\pi
\approx 0.637$. The varying force averages $2/\pi$ of its peak
value over the displacement; physically it ramps up from zero,
peaks at $x = L/2$, and ramps back to zero, so its work is the
peak-time-displacement area under a half-sine, which is
$(2/\pi)$ of the rectangle.

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Prove the linearity of the Riemann integral from the definition
in section 6.1, by passing to the limit in the linearity of the
finite Riemann sums.
\end{exercise}

\paragraph{Solution sketch.}
Fix $\varepsilon > 0$. For any partition $P$ of $[a, b]$ and
sample points $\xi_{i}$, the Riemann sum for $\alpha f + \beta g$
is $\sum_{i}(\alpha f(\xi_{i}) + \beta g(\xi_{i})) \Delta x_{i} =
\alpha \sum_{i} f(\xi_{i})\Delta x_{i} + \beta \sum_{i}
g(\xi_{i})\Delta x_{i}$. Choose $\delta > 0$ small enough that
both Riemann sums approximate their integrals within
$\varepsilon / (2 \max(\abs{\alpha}, \abs{\beta}, 1))$ for every
partition with mesh below $\delta$. The triangle inequality then
gives the combined sum within $\varepsilon$ of $\alpha \int f +
\beta \int g$, which is the definition of $\int (\alpha f + \beta
g) = \alpha \int f + \beta \int g$.

\begin{exercise}
Prove the additive property of intervals
$\int_{a}^{b} f = \int_{a}^{c} f + \int_{c}^{b} f$ from the
definition, for $f$ continuous on $[a, b]$ and $c \in (a, b)$.
\end{exercise}

\paragraph{Solution sketch.}
For any partition of $[a, b]$ that includes $c$ as a node, the
Riemann sum splits into one sum over $[a, c]$ and one over
$[c, b]$. Continuity of $f$ on $[a, b]$ implies integrability on
both subintervals (section 6.1) and on the whole. Refining the
partition to include $c$ does not change the limiting value of
the Riemann sum (the convergence is independent of the partition
choice). Passing to the limit, the left side is $\int_{a}^{b}f$,
the right is $\int_{a}^{c}f + \int_{c}^{b}f$.

\begin{exercise}
Derive the substitution rule for definite integrals from the
chain rule. State the conditions on $g$ and $f$ explicitly and
identify the step at which the limits transform.
\end{exercise}

\paragraph{Solution sketch.}
Assume $g$ is continuously differentiable on $[a, b]$ and $f$ is
continuous on $g([a, b])$. Let $F$ be an antiderivative of $f$ on
that range. By the chain rule, $\dv{}{x}\bigl(F(g(x))\bigr) =
f(g(x))\,g'(x)$. By Part II of the fundamental theorem applied
to the left side, $\int_{a}^{b} f(g(x))\,g'(x)\dd x = F(g(b)) -
F(g(a))$. By Part II applied to $f$ on $[g(a), g(b)]$, the right
side equals $\int_{g(a)}^{g(b)} f(u)\dd u$. The limits transform
in the second application of Part II: the integral over $u$ runs
between $g(a)$ and $g(b)$ because $F$ is being evaluated at those
points.

\begin{exercise}
Derive the integration-by-parts formula from the product rule of
differentiation. State the conditions on $u$ and $v$ that justify
the integration on both sides.
\end{exercise}

\paragraph{Solution sketch.}
Assume $u$ and $v$ are continuously differentiable on $[a, b]$.
The product rule says $(uv)' = u'v + uv'$. Both sides are
continuous, so integrate on $[a, b]$ by Part II of the fundamental
theorem: $\int_{a}^{b}(uv)' \dd x = [uv]_{a}^{b} = u(b)v(b) -
u(a)v(a)$. Rearranging, $\int_{a}^{b} u v' \dd x = [uv]_{a}^{b}
- \int_{a}^{b} u' v \dd x$, which is the integration-by-parts
formula. The indefinite form follows by treating $a$ as a
boundary constant absorbed into $C$.

\begin{exercise}
Derive the trapezoidal-rule error formula
$\int_{a}^{b} f - T_{n} = -(b - a) h^{2} f''(\xi) / 12$ on a single
subinterval first, by a Taylor expansion of $f$ about the midpoint
of the subinterval, and then sum.
\end{exercise}

\paragraph{Solution sketch.}
On a subinterval $[x_{i-1}, x_{i}]$ of width $h$, let $m_{i} =
(x_{i-1} + x_{i})/2$. Expand $f$ as $f(x) = f(m_{i}) +
f'(m_{i})(x - m_{i}) + \tfrac{1}{2}f''(m_{i})(x - m_{i})^{2} +
O((x - m_{i})^{3})$. Integrate over $[x_{i-1}, x_{i}]$: the
linear term integrates to zero by symmetry, and $\int(x - m_{i})
^{2}\dd x = h^{3}/12$. The trapezoidal estimate on this panel is
$h(f(x_{i-1}) + f(x_{i}))/2 = h f(m_{i}) + h \cdot \tfrac{1}{2}
f''(m_{i})(h/2)^{2} + O(h^{5}) = h f(m_{i}) + h^{3}f''(m_{i})/8 +
O(h^{5})$. The exact integral on the panel is $h f(m_{i}) +
h^{3}f''(m_{i})/24 + O(h^{5})$. The panel error is therefore
$-h^{3}f''(m_{i})/12 + O(h^{5})$. Summing over $n$ panels and
applying the mean value theorem to the sum gives
$-(b-a)h^{2}f''(\xi)/12$.

\begin{exercise}
Derive the parallel-axis theorem
$I_{\mathrm{axis}} = I_{\mathrm{cm}} + M d^{2}$ from the definition
of the moment of inertia as an integral over the mass distribution.
State the assumption (the chosen axis is parallel to the centre-of-
mass axis).
\end{exercise}

\paragraph{Solution sketch.}
Place coordinates so the centre-of-mass axis is the $z$-axis and
the parallel axis is offset by $d$ along $x$. For a point at
$(x, y, z)$ with mass element $\dd m$, the distance to the
centre-of-mass axis is $\sqrt{x^{2} + y^{2}}$ and to the parallel
axis is $\sqrt{(x - d)^{2} + y^{2}}$. So $I_{\mathrm{axis}} =
\int\bigl((x - d)^{2} + y^{2}\bigr) \dd m = \int(x^{2} + y^{2})
\dd m - 2d \int x \dd m + d^{2}\int \dd m = I_{\mathrm{cm}} -
2d \cdot M \bar{x} + d^{2} M$. The centre-of-mass coordinate
$\bar{x}$ vanishes by the definition of the centre of mass, so
$I_{\mathrm{axis}} = I_{\mathrm{cm}} + M d^{2}$.

\begin{exercise}
Derive the convolution-commutativity identity $f * g = g * f$ from
the definition $(f * g)(t) = \int_{-\infty}^{\infty} f(\tau) g(t -
\tau) \dd \tau$ by the substitution $\sigma = t - \tau$. State the
condition on $f$ and $g$ under which the manipulation of limits is
justified (absolute convergence of the integral over $\R$).
\end{exercise}

\paragraph{Solution sketch.}
With $\sigma = t - \tau$, $\dd \sigma = -\dd \tau$. When $\tau =
-\infty$, $\sigma = +\infty$; when $\tau = +\infty$, $\sigma =
-\infty$. The minus sign on $\dd \sigma$ swaps the limits back:
$\int_{-\infty}^{\infty} f(\tau) g(t - \tau) \dd \tau =
\int_{-\infty}^{\infty} f(t - \sigma) g(\sigma) \dd \sigma$,
which is $(g * f)(t)$. The manipulation is justified when the
integral converges absolutely (which lets the change of variables
proceed without conditional-convergence subtleties); $f, g \in
L^{1}(\R)$ is the standard sufficient condition.

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the total electrical energy a $60\,\si{\watt}$ lamp consumes
if left on continuously for a calendar year, in kilowatt-hours.
State your assumption about the duty cycle (continuous, or with
some fraction of the year off). Provide a factor-of-three
confidence range.
\end{exercise}

\paragraph{Solution sketch.}
A year has about $8.76 \times 10^{3}\,\text{h}$. Continuous duty:
$60\,\si{\watt} \times 8760\,\text{h} = 5.26 \times 10^{5}\,
\text{Wh} \approx 525\,\text{kWh}$. A more realistic typical
household lamp runs perhaps $4\,\text{h}/\text{day}$, giving
$60 \times 1460 = 87.6\,\text{kWh}$. The factor-of-three range
covers anything from a closet lamp seldom used ($\sim 50\,
\text{kWh}$) to a continuously lit security lamp ($\sim 525\,
\text{kWh}$). At a 2026 residential rate of about
\$0.20$/\text{kWh}$, the continuous-duty cost is about \$105/yr.

\begin{exercise}
Estimate the integral $\int_{0}^{\infty} e^{-x^{2}} \dd x$ to within
a factor of two using only the comparison
$e^{-x^{2}} \leq e^{-x}$ for $x \geq 1$ and a guess for the
contribution from $[0, 1]$. State your guess and your factor-of-
two confidence claim. The exact value $\sqrt{\pi}/2$ may be quoted
afterwards as a calibration.
\end{exercise}

\paragraph{Solution sketch.}
On $[0, 1]$, $e^{-x^{2}}$ falls from $1$ to $1/e \approx 0.37$.
Linear interpolation gives an average of about $0.68$, so the
contribution to the integral is about $0.68$. On $[1, \infty)$,
the comparison $e^{-x^{2}} \leq e^{-x}$ gives an upper bound
$\int_{1}^{\infty} e^{-x} \dd x = 1/e \approx 0.37$; the actual
contribution is somewhat smaller (because the Gaussian falls
faster), say $0.15$ as a guess. The estimate is then $0.68 + 0.15
= 0.83$; the actual value is $\sqrt{\pi}/2 \approx 0.886$, so the
estimate is within $7\,\%$, comfortably inside the factor-of-two
claim.

\begin{exercise}
Estimate the moment of inertia of a hardback book about an axis
through its spine, by treating the book as a uniform rectangular
prism. State your assumed dimensions, mass, and the axis location.
Quote a factor-of-three confidence range and state which dimension
or mass uncertainty dominates.
\end{exercise}

\paragraph{Solution sketch.}
A representative hardback: $24\,\text{cm}$ tall, $16\,\text{cm}$
wide, $3\,\text{cm}$ thick, mass $0.9\,\text{kg}$. The axis runs
through the spine, parallel to the height; the moment is then
$I = (M/12)(w^{2} + t^{2}) = (0.9/12)(0.16^{2} + 0.03^{2}) =
0.075 \cdot (0.0256 + 0.0009) = 0.075 \cdot 0.0265 \approx 2.0
\times 10^{-3}\,\text{kg m}^{2}$. The width term dominates by a
factor of $\sim 30$, so width uncertainty (which is $\sim 10\,\%$
across hardback formats) governs the result. A factor-of-three
range covers everything from a small novella ($\sim 5 \times 10
^{-4}$) to an oversized art book ($\sim 6 \times 10^{-3}$).

\begin{exercise}
Estimate the work done in lifting a $10\,\si{\kilogram}$ bag of
sand from the floor to a shelf $1.5\,\si{\meter}$ above the floor.
State the integration that turns the variable height into the work
calculation and identify the assumption that lets the answer
collapse to $m g h$.
\end{exercise}

\paragraph{Solution sketch.}
The work is $W = \int_{0}^{h} F(z) \dd z$ where $F(z)$ is the
upward force required to hold the bag at height $z$. For a rigid
bag in a uniform gravitational field, $F(z) = m g$ is constant,
and the integral collapses to $W = m g h = 10 \cdot 9.81 \cdot
1.5 \approx 147\,\text{J}$. The collapsing assumption is twofold:
the bag is rigid (so all its mass moves the same distance) and
$g$ is constant over the lifting range (true to better than
$10^{-6}$ at $1.5\,\text{m}$).

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement the trapezoidal rule and Simpson's rule for
$\int_{0}^{1} e^{-x^{2}} \dd x$ in your preferred programming
environment. Plot the absolute error of each rule against the
number of subintervals $n$ on a log-log scale, for $n$ ranging
from $4$ to $1024$. Verify the slopes match the theoretical
$h^{2}$ and $h^{4}$ scaling.
\end{exercise}

\paragraph{Solution sketch.}
A reference implementation is in
\texttt{code/quadrature\_convergence.py}. The exact value is
$(\sqrt{\pi}/2)\,\mathrm{erf}(1) \approx 0.7468241328$. The
expected output: trapezoidal error falls by a factor of $4$ each
doubling of $n$ (slope $-2$ on log-log), Simpson by a factor of
$16$ (slope $-4$), until floating-point cancellation in the
weighted sum dominates somewhere around $n = 10^{5}$. The plot
should show two straight lines of the predicted slopes, joining
the data points at small $n$ where rounding error is invisible.

\begin{exercise}
Implement three-point Gauss-Legendre quadrature for
$\int_{0}^{\pi/2} \sqrt{\sin x} \dd x$. Compare the result to a
trapezoidal rule with $n = 100$ subintervals. State the function-
evaluation count for each and identify which method gives more
accuracy per evaluation.
\end{exercise}

\paragraph{Solution sketch.}
The reference value (via a high-order rule) is
$\int_{0}^{\pi/2}\sqrt{\sin x} \dd x \approx 1.19814$. Three-point
Gauss-Legendre on $[0, \pi/2]$ with $3$ evaluations gives about
$1.196$; relative error $\sim 2 \times 10^{-3}$. The integrand
has an integrable singularity in its derivative at $x = 0$
($\sqrt{\sin x} \sim \sqrt{x}$), so neither rule converges at its
nominal asymptotic rate. The trapezoidal rule with $n = 100$
gives about $1.197$; relative error $\sim 8 \times 10^{-4}$. The
trapezoidal rule wins on absolute accuracy because it has $101$
evaluations versus $3$; per evaluation Gauss-Legendre is far
ahead. A composite three-point Gauss-Legendre with $33$ panels
($99$ evaluations) brings the Gauss result below $10^{-5}$,
restoring the per-evaluation advantage.

\begin{exercise}
Implement the discrete convolution
$(f * g)[n] = \sum_{k} f[k] g[n - k]$ for two finite signals of
length $N = 100$ each, and compare its output to the result of a
direct trapezoidal-rule evaluation of the convolution integral on
a sampled-signal interpretation of $f$ and $g$. Plot both outputs
and identify the boundary effects at the start and end of the
output.
\end{exercise}

\paragraph{Solution sketch.}
A reference implementation is in
\texttt{code/rc\_convolution.py}; the same NumPy
\texttt{np.convolve} call with \texttt{mode="full"} times the
sample interval $\dd t$ gives the discrete approximation to
$(f * g)(t)$. The two outputs agree to within the trapezoidal
discretisation error in the interior; at the start (the first
$\sim N/2$ samples) the discrete convolution treats samples
outside the recorded interval as zero, producing a ramping-on
artefact. The remedy is to record a margin of zero-padding on
both sides or to use \texttt{mode="valid"} and report only the
unambiguous central portion.

\begin{exercise}
Implement a Monte Carlo integration of $\int_{0}^{1}\int_{0}^{1}
e^{-(x^{2} + y^{2})} \dd x \dd y$ with $N = 10^{4}$ samples. Report
the mean and standard error of the estimate. Compare to a
trapezoidal-rule double integral on a $40 \times 40$ grid.
\end{exercise}

\paragraph{Solution sketch.}
A reference implementation is in
\texttt{code/monte\_carlo\_2d.py}. The exact value is about
$0.5577$. Monte Carlo with $N = 10^{4}$ gives an estimate with
standard error $\sigma_{f}/\sqrt{N}$ where $\sigma_{f} \approx
0.18$ is the standard deviation of the integrand over the unit
square; the standard error is therefore $\sim 1.8 \times 10^{-3}$.
The $40 \times 40$ trapezoidal grid gives an error $\sim 10^{-4}$
on the smooth integrand, comfortably below the Monte Carlo
standard error and at $\sim 1600$ evaluations versus $10000$.
Monte Carlo wins in higher dimensions where the trapezoidal grid
grows as $n^{d}$.

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a quadrature procedure for an integrand that is smooth on
$(0, 1]$ but has an integrable singularity like $1/\sqrt{x}$ at
$x = 0$. State the change of variables, the resulting integrand,
and the rule applied. Justify your choice over a naive Simpson's
rule applied to the original integrand.
\end{exercise}

\paragraph{Discussion.}
Substitute $x = u^{2}$, $\dd x = 2 u \dd u$, so $\int_{0}^{1}
f(x)/\sqrt{x}\,\dd x = 2 \int_{0}^{1} f(u^{2}) \dd u$. The
transformed integrand is smooth on $[0, 1]$ if $f$ is, and
Simpson's rule (or three-point Gauss-Legendre per panel) attains
its nominal order of convergence on the transformed integrand. A
naive Simpson on the original integrand misbehaves because the
fourth derivative of $f(x)/\sqrt{x}$ blows up at the origin: the
error formula's $\xi$-dependent constant becomes unbounded, and
no fixed-step rule converges at the predicted rate. The change of
variables is the cheaper fix; adaptive quadrature with
singularity detection is the more general alternative when the
singularity's form is unknown.

\begin{exercise}
Design an experimental procedure for measuring the moment of
inertia of a rolling object (a soup can, a marble) by rolling it
down an inclined plane and measuring the time to roll a known
distance. State the integration relation between the moment of
inertia and the rolling acceleration, and the dominant
uncertainty source.
\end{exercise}

\paragraph{Discussion.}
For an object of mass $m$ and radius $R$ rolling without
slipping down a plane inclined at angle $\theta$, the
translational acceleration is $a = g \sin\theta / (1 + I/(mR^{2}))$.
Solving, $I = m R^{2}(g \sin\theta / a - 1)$. Measure $a$ by
timing the roll over a known distance $d$ from rest:
$a = 2 d / t^{2}$. The dominant uncertainty is usually the timing
of the start ($\sim 50\,\text{ms}$ on a stopwatch, much less with
a photogate) and the rolling-friction departure from the
ideal-no-slip assumption (an error of order a few per cent that
biases $I$ low). A second incline at a different angle gives an
internal consistency check; agreement between angles within their
combined uncertainties validates the no-slip assumption.

\begin{exercise}
Design a sampling plan for measuring the cross-sections of a
household object whose moment of inertia is to be computed by
numerical integration. State the sampling interval, the
measurement instrument and its resolution, and the expected
expanded uncertainty of the resulting moment of inertia. Justify
your choice of sampling interval against the convergence-by-
halving criterion.
\end{exercise}

\paragraph{Discussion.}
For an object of length $L \sim 30\,\text{cm}$ (typical hammer or
bottle), measure cross-sections at $5\,\text{mm}$ axial intervals
($\sim 60$ samples). Use a digital calliper with $0.05\,\text{mm}$
resolution; each cross-section area carries $\sim 2\,\%$ relative
uncertainty if the section is approximated as an ellipse from two
semi-axes. The integral $I = \int x^{2} \sigma(x) \dd x$ inherits
about $\sim 3\,\%$ from the measurement chain. Halving the
sampling to $2.5\,\text{mm}$ on a representative segment and
comparing should change the integral by less than the
$3\,\%$ measurement floor; if it changes more, the $5\,\text{mm}$
sampling was too coarse and the working interval is $2.5\,\text{mm}$.

\begin{exercise}
Design a procedure for verifying the convergence of a numerical
integration of a measured signal whose true integral is unknown.
State the convergence criterion, the refinement strategy, and the
stopping rule.
\end{exercise}

\paragraph{Discussion.}
Compute the integral at sampling interval $h$, $h/2$, $h/4$, and
so on (by subdividing existing samples through interpolation, or
by collecting new samples at the finer interval if the signal can
be re-acquired). The convergence criterion is $\abs{I_{h/2} -
I_{h}} < \mathrm{tol}$, where the tolerance is set by the
downstream use. The Richardson extrapolation step
$I^{*} = I_{h/2} + (I_{h/2} - I_{h})/(2^{p} - 1)$, with $p$ the
rule's order, gives a higher-order estimate whose own convergence
check ($\abs{I_{h/4}^{*} - I_{h/2}^{*}}$) is a more sensitive
proxy. Stopping rule: terminate when two successive halvings
agree within the tolerance; if they fail to agree as $h$ shrinks
indefinitely, the integrand is too irregular for the chosen rule
or the data carries noise the rule cannot integrate.

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A trapezoidal-rule integration of a sampled signal returns a value
that depends strongly on the choice of sampling phase (where the
samples land relative to the signal's features). State three
plausible mechanisms and propose the next test that distinguishes
them.
\end{exercise}

\paragraph{Solution sketch.}
Three mechanisms. First, the integrand has high-frequency content
above the sampling Nyquist limit, and aliasing makes the rule
read different totals depending on phase. Test: low-pass filter
the signal at half the sample rate and re-integrate; if the
result becomes phase-independent, aliasing was the cause. Second,
the integrand has sharp features (spikes, kinks) whose sample
coverage is phase-dependent; integrating with finer sampling
across those features removes the phase dependence. Third, the
trapezoidal rule's error formula assumes a twice-differentiable
integrand and the signal is too irregular; the test is to switch
to a rule designed for irregular signals (a midpoint rule, an
adaptive rule with refinement on the irregular regions).

\begin{exercise}
A numerical integration of $\int_{0}^{1} 1/\sqrt{x} \dd x = 2$
returns a value of $1.7$ from a Simpson rule with $n = 10$
subintervals, and $1.85$ from $n = 100$. State the failure mode
and propose a corrected procedure.
\end{exercise}

\paragraph{Solution sketch.}
The integrand has a singularity at $x = 0$, so Simpson's rule
underestimates the contribution from the near-origin region.
Refining from $n = 10$ to $n = 100$ improves the answer by $0.15$
but does not recover the true value because the singularity is
not removed by uniform refinement. The corrected procedure is the
substitution $x = u^{2}$, $\dd x = 2 u \dd u$, giving
$\int_{0}^{1} 2 \dd u = 2$ exactly. Alternatively, an adaptive
rule with finer panels near $x = 0$ converges, but only at the
cost of many panels.

\begin{exercise}
A Monte Carlo estimate of an expected value of a random variable
returns a sample mean that does not stabilise as the sample size
grows: every doubling of the sample size produces a substantially
different estimate. State the diagnostic that distinguishes a
heavy-tailed distribution (where the expectation diverges) from a
finite-expectation distribution sampled with a bug.
\end{exercise}

\paragraph{Solution sketch.}
Plot the running mean against the sample size on a log $x$-axis.
A finite-expectation distribution with high variance shows a
random walk that narrows as $1/\sqrt{N}$; a heavy-tailed
distribution shows occasional large jumps that grow with $N$ (the
maximum observation grows at a rate determined by the tail
exponent). Plot also the maximum observation against $N$: for a
distribution with finite expectation, $\max / N \to 0$; for a
heavy tail with $\alpha < 1$, the maximum grows faster than $N$
and the sample mean is dominated by the maximum. A bug in the
sampler typically produces a deterministic drift or a structured
non-stationarity that direct inspection of the sample sequence
reveals.

\begin{exercise}
A reader computes the moment of inertia of a hammer about an axis
through its head by analytical decomposition (head plus handle,
each as a simple shape) and gets $I_{1}$. The reader then computes
the same moment of inertia by numerical integration on a measured
profile and gets $I_{2} = 1.5 I_{1}$. State two plausible
mechanisms for the discrepancy and propose, for each, the test
that would confirm or refute it.
\end{exercise}

\paragraph{Solution sketch.}
First mechanism: the analytical decomposition under-counts the
handle's contribution to the moment about the head end, because
the handle's far end is at $\sim L$ from the axis and contributes
$\sim m_{\mathrm{handle}} L^{2}/3$. If the handle's mass or length
was underestimated, $I_{1}$ is biased low. Test: weigh the
handle and head separately (cut a comparable hammer, or estimate
from the head's volume and the handle's density) and recompute
$I_{1}$ with the correct masses. Second mechanism: the
numerical-integration profile has an axial-origin error so the
$x^{2}$ weighting integrates against shifted coordinates,
overstating the moment. Test: vary the assumed origin in the
profile script by $\pm 5\,\text{mm}$; if $I_{2}$ moves by $\sim
30\,\%$, the origin error is responsible.

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A signal-processing engineer integrates a sinusoid sampled at
exactly twice the signal frequency (the Nyquist rate) by the
trapezoidal rule. Identify the failure mode and the sampling
condition under which the integration becomes reliable. Reference
the section 6.8 discussion of high-frequency oscillation at the
sampling scale.
\end{exercise}

\paragraph{Solution sketch.}
At exactly the Nyquist rate (two samples per period), the
trapezoidal rule returns a value that depends entirely on the
phase: samples at the peaks give a large sum, samples at the
zero crossings give zero. The true integral of $\sin(\omega t)$
over an integer number of periods is zero. The rule becomes
reliable only well above the Nyquist rate; engineering practice
oversamples by a factor of three to ten so that the trapezoidal
panels span well below one period each. Below Nyquist (less than
two samples per period) the integration is hopelessly aliased and
no rule on the existing samples can recover the true integral.

\begin{exercise}
A risk analyst computes the expected value of a Pareto-distributed
loss with shape parameter $\alpha = 0.8$. State the failure mode
and identify the threshold value of $\alpha$ above which the
expectation exists. Discuss whether the median, the truncated
mean, or another summary statistic is the appropriate report
when the expectation does not exist.
\end{exercise}

\paragraph{Solution sketch.}
The Pareto density $f(x) = \alpha x_{\min}^{\alpha} / x^{\alpha +
1}$ for $x \geq x_{\min}$ has expectation
$\int x f(x) \dd x = \alpha x_{\min}^{\alpha} \int x^{-\alpha} \dd
x$, which converges at infinity only for $\alpha > 1$. At $\alpha
= 0.8$ the integral diverges; the sample mean does not stabilise
and the reported expected value is a number that depends on the
sample, not on the distribution. The threshold is $\alpha > 1$
for the mean, $\alpha > 2$ for the variance. When the
expectation does not exist, the median is the only well-defined
location statistic; a truncated mean (with an explicit truncation
threshold and a sensitivity analysis on the threshold) is a
working compromise that the analyst must report transparently as
truncated, not as ``the mean.''

\begin{exercise}
An automated scientific instrument integrates a measured spectrum
over a wavelength range using a fixed-step trapezoidal rule, and
reports a total flux. The instrument's data-acquisition system
introduces narrow electrical-pickup spikes whose width is shorter
than the sampling interval. State whether the reported flux is
biased high, low, or unbiased, and propose a defence against the
mechanism.
\end{exercise}

\paragraph{Solution sketch.}
The bias depends on the spike rate per sample and the spike's
sign. If the spikes are positive and occur with rate $r$ samples
per spike on average, the average sample carries an extra
contribution of (spike amplitude $\times$ duration $/$ sample
interval). The trapezoidal rule sees these as genuine signal and
the reported flux is biased high by that amount. If spikes are
narrow enough to be missed by some samples and caught by others,
the integrated bias is the spike amplitude times spike duration
times spike rate, regardless of sampling phase, because in the
aggregate every spike contributes some product of amplitude and
duration. The defence is a notch filter or median filter applied
to the raw samples before integration, or oversampling with a
boxcar averager that integrates the spikes deterministically.

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
Argue, in 200 to 300 words, whether a working engineer should
prefer Simpson's rule or three-point Gauss-Legendre quadrature for
integrating a smooth function known analytically. Take a position;
defend it with a specific scenario in which your preferred answer
fails.
\end{exercise}

\paragraph{Discussion.}
Defensible position: prefer Gauss-Legendre because for smooth
integrands its polynomial-exactness gains beat Simpson's
sample-by-sample. Three-point Gauss-Legendre matches Simpson's
$n = 6$ polynomial exactness with half the function evaluations,
and a composite eight-panel three-point Gauss matches
$n \approx 100$ Simpson at $24$ evaluations. The defensible
failure mode is integrand discontinuity: a piecewise definition
with a jump in the middle of a Gauss panel violates the
polynomial-exactness premise, and Gauss errs by a fraction of the
jump per panel regardless of panel count, while Simpson with the
jump on a node boundary returns the correct sum of two integrals.
Defending the alternative (Simpson) position rests on the cost of
inserting Gauss tables and the awkwardness of non-equispaced
nodes when the integrand is sampled on a fixed grid, which in
engineering practice is the usual case. The reader who picks
either position is on defensible ground; the AI-generic answer
``it depends'' is not, because every working engineer must, in
fact, pick on a given problem.

\begin{exercise}
A senior colleague tells the reader that ``numerical integration
is an implementation detail; the analytical answer is the truth.''
Write a one-paragraph reply that takes the colleague seriously,
identifies the regime in which the claim is defensible, and
identifies the regime in which it is not.
\end{exercise}

\paragraph{Discussion.}
The colleague is defensible where the analytical answer exists,
is in a closed form whose value can be evaluated to higher
precision than the engineering use requires, and the numerical
rule is a backup check against the analytic derivation having an
algebra error. The colleague is wrong in two regimes. Most
engineering integrands of practical importance (measured spectra,
finite-element residuals, Monte Carlo expectations, convolutions
with measured impulse responses) have no analytical closed form;
the numerical rule is the answer and its convergence study is
the analyst's commitment to the answer's accuracy. Second,
analytical formulas evaluated at scale on floating-point hardware
themselves carry numerical error from elementary-function library
implementations, ill-conditioned subtractions, and overflow; the
analytical formula is then not the truth but a different
numerical computation whose error is hidden by being implicit
rather than declared.

\begin{exercise}
Argue, in 200 to 300 words, whether the convolution integral or
the Fourier transform is the more useful representation of an LTI
system for an engineer who must reason about the system's
behaviour without computational aids. Take a position and defend
it with reference to a working engineering domain (audio, control,
imaging, or another).
\end{exercise}

\paragraph{Discussion.}
Defensible position: the convolution integral is more useful for
direct intuition about time-domain behaviour, and the Fourier
transform is more useful for frequency-domain reasoning. The
engineer in audio reasoning about a reverb knows the impulse
response of the room as a measurable artefact and the convolution
gives the time-domain output directly; the Fourier transform of
that impulse response, while equivalent, is not what an audio
engineer's ear tracks. The engineer in control theory reasoning
about stability of a closed loop wants the Fourier (or Laplace)
description because pole locations in the complex plane carry the
stability information directly; the convolution that produces
the step response gives information about the same poles but
through a transient that the engineer must read backwards to
recover them. The defensible position is that the choice depends
on the question; the working compromise is fluency in both, with
the discipline of using the simpler representation for the
problem at hand rather than the more familiar one.

\begin{exercise}
Argue, in 200 to 300 words, whether the engineering working rule
``use Simpson on smooth data, trapezoidal on noisy data, Gauss on
analytical integrands'' is a defensible heuristic or a habit that
hides the working trade. Take a position and defend it with a
specific case where the heuristic gives the wrong recommendation.
\end{exercise}

\paragraph{Discussion.}
The heuristic is defensible as a starting position for an
engineer with no other information about the integrand, because
each rule is appropriate to the integrand class the rule's error
formula assumes. The heuristic hides the working trade in three
common cases. First, ``noisy data'' covers both random
measurement noise (where trapezoidal is honest) and structured
artefacts like power-line pickup or quantisation noise (where
filtering then Simpson is better than trapezoidal on the raw
data). Second, ``smooth data'' covers both genuinely smooth
samples and over-sampled smooth signals where Simpson's
higher-order error formula has not yet caught up with
trapezoidal's lower-constant error; for $n$ in the low tens, the
two rules can give comparable accuracy on the same data, and the
choice rests on implementation cost. Third, ``analytical
integrands'' includes integrands with hidden singularities or
high-frequency content that defeats Gauss-Legendre's polynomial
exactness assumption. The defensible response is to take the
heuristic as the default and to make the integrand's class
explicit before defaulting; an analyst who applies the rule
without that check is making the heuristic do work it cannot
support.
