\chapter{Design reviews, certification, sign-off}
\label{vol11:ch09}

\epigraph{The signature commits the engineer to the design's adequacy; the design review is the discipline by which the signature is earned rather than affixed.}{}

\chapmeta{Half-life: medium-life. Archetypes: ethics, failure. Exercise target: 15. Project track: hybrid.}

\noindent
The design review is the engineering practice by which a design's adequacy is assessed by qualified parties before the design proceeds to its next stage or to release. Certification is the regulatory equivalent: the assessment by which a design is approved for the regulated activity. Sign-off is the commitment a specific engineer makes to the design's adequacy through their signature on the design documents. The three are connected: the review produces the evidence; the certification is the formal approval; the sign-off is the individual engineer's commitment. This chapter develops the practice, the structures, the failure modes (rubber-stamp reviews; certification gaps; uncritical sign-off), and the professional and legal implications.

\section{The design review as engineering practice}\pathtag{core}

A design review is a structured assessment of a design by qualified parties who are not the design's author. The purpose is to identify issues the design team has missed and to confirm the design's adequacy for proceeding. The review's value depends on its substantive engagement; reviews that confirm without scrutiny produce theatre rather than evidence.

The activities of a substantive review:

\textbf{Pre-review preparation.} The reviewers receive the design documentation in advance: the requirements; the design itself (drawings, analyses, test results); the rationale for design decisions; the open issues. The reviewers study the materials and prepare specific questions and concerns; arriving at the review unprepared produces shallow engagement.

\textbf{The review session.} The design team presents the design; the reviewers question, challenge, and identify concerns; the team responds; concerns that cannot be resolved in the session are recorded for follow-up. The session is structured: a defined agenda; defined roles; defined documentation.

\textbf{Concerns and action items.} Each concern raised in the review is documented: the concern; the responsible party; the proposed resolution; the timeline. Concerns are tracked through resolution; the review does not close until all concerns are addressed (resolved; deferred with rationale; or escalated).

\textbf{The review record.} The review's proceedings, concerns, and resolutions are documented persistently. The record is part of the design's history; it supports subsequent reviews, certification submissions, and forensic analysis if the design fails in service.

\textbf{The pass/fail decision.} The review reaches a decision: the design proceeds; the design proceeds with conditions (specific actions required before the next stage); the design returns for revision. The decision is the review's substantive output.

The functions a design review serves:

\textbf{Catches the design team's blind spots.} Every design team has blind spots: assumptions that were unstated; alternatives that were not considered; failure modes that were not anticipated. The reviewers, looking at the design with fresh eyes, surface what the team did not see.

\textbf{Brings external expertise.} The reviewers may have expertise the design team lacks: domain experience in similar designs; analytical skills in particular techniques; knowledge of standards and regulations the team is not familiar with. The expertise is contributed at the review.

\textbf{Aligns stakeholders.} The review brings the parties whose subsequent work depends on the design (manufacturing; service; safety; certification) into early engagement with the design. Issues that would otherwise surface at handoff are surfaced earlier.

\textbf{Tests the design team's understanding.} The review questions the design team's reasoning; the team's ability to explain and defend the design tests their understanding. Gaps in understanding surface as gaps in the explanation.

\textbf{Provides the engineering organisation's institutional memory.} The review's record becomes part of the design's history; it supports the institution's learning across designs and across generations.

The discipline of the design review is the discipline of producing substantive engagement rather than ceremony. The discussion below addresses what produces substantive reviews and what produces rubber-stamping.

\begin{principle}[The review is the discipline of external scrutiny]
A design review's value is the external scrutiny it brings to bear on the design team's work. Reviews that consist of design-team presentation followed by polite acknowledgement produce no engineering value; reviews in which the reviewers ask substantive questions and the design team's responses are recorded produce evidence the design has been challenged. The discipline of the engineering profession is the discipline of demanding the substantive form.
\end{principle}

\section{PDR, CDR, FDR}\pathtag{core}

The dominant formal-review structure across many engineering disciplines uses three review stages: Preliminary Design Review (PDR), Critical Design Review (CDR), and Final Design Review (FDR; sometimes called Production Readiness Review or PRR).

\textbf{Preliminary Design Review (PDR).} Conducted after concept selection (Chapter 4) and before detailed design (Chapter 5). The PDR reviews: the selected concept against requirements; the design approach (architecture, technology choices, key tradeoffs); the preliminary analyses (concept-level structural, thermal, performance estimates); the risks identified and their mitigation plans; the schedule and resources for detailed design. The PDR's pass criterion is that the concept is sound and adequately developed for detailed design to proceed.

\textbf{Critical Design Review (CDR).} Conducted after detailed design (Chapter 5) and before commitment to manufacturing or construction. The CDR reviews: the detailed design (drawings, models, specifications); the detailed analyses (FEA, thermal, electrical, performance); the verification plan and any verification already completed; the prototype evidence; the manufacturing readiness; the production cost and schedule. The CDR's pass criterion is that the detailed design is verified to meet requirements and is ready for manufacturing commitment.

\textbf{Final Design Review (FDR) / Production Readiness Review (PRR).} Conducted before production release or before service introduction. The FDR reviews: the design's complete verification evidence; the manufacturing process qualification; the pilot production results; the certification status; the service-support readiness; the release plan. The FDR's pass criterion is that the design is ready for release.

Additional reviews appear in various traditions: System Requirements Review (SRR) at the start; Preliminary Requirements Review (PRR-equivalent); Test Readiness Review (TRR) before each major test campaign; Operational Readiness Review (ORR) before first operational use. The specific names and exact gates vary by industry and organisation.

The review structure has several functions:

\textbf{Stage-gates against premature commitment.} Each review is a checkpoint at which the design's adequacy for the next stage is assessed; designs that have not adequately developed cannot proceed without explicit acceptance of the residual risk.

\textbf{Evidence accumulation.} Each review's evidence accumulates: by the FDR, the design's complete history is documented and reviewed. The accumulation supports certification (which often requires complete review records) and forensic analysis if the design later fails.

\textbf{Schedule discipline.} The reviews are scheduled events; the project's work organises against them. The schedule discipline supports project planning and resource allocation.

\textbf{Stakeholder engagement.} The reviews are the occasions at which the broader engineering community engages with the design: manufacturing reviewers attend the CDR; certification engineers attend throughout; the customer's technical representatives may attend specific reviews. The engagement is continuous through the design process rather than concentrated at handoff.

The discipline of stage-gated reviews is documented in many standards: ISO/IEC/IEEE 15288 (systems engineering); various NASA, DoD, and ESA standards; the major industry-specific frameworks (AS9100 for aerospace; ISO 13485 for medical devices; ASME boiler and pressure vessel code for pressure equipment). The exact gates and content differ; the underlying discipline is similar.

\section{Independent review and adversarial design}\pathtag{standard}

The reviews of the previous section are typically conducted by reviewers who are part of the same organisation as the design team. The structural risk is that the reviewers share the design team's frames, assumptions, and blind spots; reviews that should be substantive can become collegial discussions that do not surface the issues the design team did not see.

Independent review addresses the risk by introducing reviewers from outside the organisation: independent engineering consultants; academic reviewers; sister-organisation engineers; regulator-employed technical staff. The independence brings perspective that internal review does not.

The adversarial-design technique extends independent review further: an adversarial team is tasked with attacking the design, identifying its weakness, finding the conditions under which it fails. The adversarial team's success is measured by what they find; the design team's response is the design's improvement.

The mechanisms:

\textbf{Red teams in security.} In security engineering, the red team's job is to attack the system as a hostile actor would. The team's findings drive defensive improvements. The technique is widely used in IT security, in physical security, and increasingly in safety analyses where adversarial framing helps surface failure conditions.

\textbf{Tiger teams in aerospace.} In aerospace, the tiger team is a group of senior engineers convened to address a specific problem with concentrated expertise. Tiger teams are often used in post-incident analysis (the Apollo 13 tiger team; the Columbia accident's mishap investigation) but can also be used pre-launch as adversarial reviewers.

\textbf{Devil's advocate in design reviews.} A specific reviewer is assigned the devil's advocate role: their job is to challenge the design's assumptions, to argue for alternatives, to surface concerns the other reviewers have not raised. The role is structurally adversarial; it produces challenges that polite review does not.

\textbf{Independent technical authority.} Some engineering organisations maintain an independent technical authority (an Independent Verification and Validation (IV\&V) organisation; a Chief Engineer's office; a Safety Office) whose job is to provide independent technical judgment on design decisions. The post-Columbia NASA reforms established this structure for the Space Shuttle and subsequent programmes \cite{acc:caib-columbia}.

\textbf{External review boards.} Some projects engage formal external review boards: blue-ribbon panels; industry-expert advisory boards; standing review committees. The boards provide periodic independent assessment; their findings inform the design organisation's leadership.

The cost of independent and adversarial review is real: the reviewers' time; the organisational disruption of substantive challenge; the resistance to findings that the design team does not want to accept. The cost of inadequate review is the failure mode that the review would have caught.

The Volume X cases include multiple instances where independent review was inadequate or absent. The Boeing 737 MAX MCAS certification was conducted under the FAA's ODA arrangement where Boeing-employed authorised representatives served as the formal reviewers; the structural independence was inadequate to the substantive challenge required \cite{acc:house-737max-2020}. The Therac-25 software was developed without independent review of the dose-control software; the race condition that produced the accidents was not surfaced by internal review \cite{paper:leveson-turner1993}. The pattern is recurring: organisations that maintain substantive independent review have better safety records; organisations that do not have worse ones.

\section{Certification regimes revisited at design depth}\pathtag{standard}

Volume X Chapter 11 developed certification regimes as institutional mechanisms. This section revisits them at design depth: what the design team must do to engage with certification, what evidence the certification requires, what the certification process looks like from the design organisation's perspective.

The dominant regimes and their design-level requirements:

\textbf{Aviation type certification (FAA, EASA).} The design organisation must demonstrate compliance with the applicable airworthiness standards (FAR Part 25 for transport aircraft; the rotorcraft and small-aircraft equivalents). The compliance is demonstrated through a combination of analysis (computational evidence), test (physical evidence), and inspection (manufacturing-process evidence). The design organisation maintains a Type Inspection Authorisation (TIA) that documents the certification basis and the compliance evidence. The certification engineering team coordinates with the FAA's Aircraft Certification Office (ACO) throughout; the engagement is structured (Type Inspection Authorisation, issue papers, and the structured process of identifying and closing certification issues). The certification process produces the Type Certificate that authorises production and operation.

\textbf{Medical device approval (FDA in the US; EMA in Europe; the various national equivalents).} The design organisation must demonstrate the device is safe and effective for its intended use. The pathway depends on the device's risk class: Class I requires minimal pre-market activity; Class II typically requires 510(k) substantial-equivalence demonstration; Class III requires Pre-Market Approval (PMA) with clinical evidence. The design organisation maintains a Design History File (DHF) that documents the design's development under the Quality System Regulation (21 CFR 820). The DHF includes: requirements; design inputs and outputs; verification and validation evidence; risk management documentation (ISO 14971); software-development documentation (IEC 62304 if software is involved); usability documentation (IEC 62366); biocompatibility evidence (ISO 10993). The FDA's review process examines the DHF and the additional submission elements (clinical evidence; labelling; manufacturing controls).

\textbf{Nuclear plant licensing (NRC).} The design organisation (for new reactor designs; the operator for plant-specific licensing) must demonstrate that the design meets the technical specifications and regulatory requirements. The design certification process for new reactor types involves the design organisation's submission of a Design Certification Document; the NRC's review is multi-year and involves multiple Advisory Committee on Reactor Safeguards (ACRS) reviews. The Plant licence is then specific to each site, with the operator demonstrating site-specific compliance.

\textbf{Pressure vessel and boiler certification (ASME BPVC).} The design organisation must demonstrate compliance with the applicable ASME Boiler and Pressure Vessel Code. The compliance is demonstrated through: design calculations to the code's formulae; material certifications; welder qualifications; inspection records. The design is stamped (literally or figuratively) by the Authorised Inspector who confirms compliance.

\textbf{Building and structural certification.} The design organisation (architecture firm; structural-engineering firm) must demonstrate compliance with the applicable building codes (International Building Code in the US; Eurocodes in Europe; national equivalents). The Professional Engineer signs and seals the structural drawings; the seal carries legal liability for the design's compliance. The Authority Having Jurisdiction (the building department; the structural-engineering reviewer) issues the building permit on the basis of the sealed documents.

\textbf{Software safety certification (DO-178C for avionics; ISO 26262 for automotive; IEC 61508 for industrial).} The design organisation must demonstrate the software meets the integrity level appropriate to its safety classification. The evidence includes: requirements; design documentation; code; test coverage; configuration management; verification of each life-cycle stage. The certification authority reviews the evidence; the certification's reliance is on the documented process more than on the software's properties directly.

The design organisation's engagement with each regime requires substantial certification engineering: a function dedicated to managing the regulatory interaction. The function's discipline includes: knowing the applicable requirements in detail; producing the evidence the requirements demand; engaging with the regulator throughout the design rather than only at submission; closing the regulator's issues as they arise.

\section{The signature: what an engineer commits when they sign}\pathtag{core}

The engineering signature on a design document commits the engineer to the design's adequacy. The commitment has legal weight; it has professional weight; it has personal weight. The signature is not a formality; it is the engineer's professional commitment that the design satisfies the relevant requirements within the scope of the engineer's responsibility.

The signature's substance:

\textbf{Professional certification.} In jurisdictions that license engineers (the US Professional Engineer (PE) licence; the UK Chartered Engineer (CEng) designation; the Canadian P.Eng.; the various national equivalents), the licensed engineer's signature carries the licence's authority. The licence is granted on the basis of education, experience, and examination; it is maintained through continuing professional education and through compliance with the licensing body's ethics code. The signature commits the licensed engineer to the design's compliance with the standards of practice the licence implies.

\textbf{Legal liability.} The engineer who signs a design is exposed to legal liability if the design fails and the failure is attributed to the engineer's lapse. The liability varies by jurisdiction and by the specifics of the engineer's employment (employee-engineer; consulting engineer; principal-of-firm); the engineer's professional indemnity insurance covers the liability up to its limits. The exposure is real; the discipline of cautious signature is part of the engineer's professional practice.

\textbf{Organisational commitment.} The engineering organisation that produces the design is committed to the design's adequacy through the engineer's signature. The organisation's quality system, its insurance, its institutional reputation are all linked to the signature.

\textbf{The customer's reliance.} The customer who receives the design relies on the signature as evidence of professional commitment. The customer's decisions (procurement; construction; operation) proceed on the basis of the signed design.

\textbf{The regulator's evidence.} The regulator who reviews the design treats the signature as evidence of professional commitment by a qualified engineer. The regulator's processes assume the signature has been thoughtfully applied.

\textbf{The next engineer's reliance.} An engineer who subsequently uses the design (in maintenance; in revision; in next-generation design) relies on the signature as evidence of the design's adequacy at the time of signing. The reliance is across time; the signature's commitment persists through the design's service life.

The discipline of substantive signature includes:

\textbf{Understanding what is being signed.} The engineer reads the design document before signing; signs only what the engineer has examined; does not sign documents the engineer does not understand or has not reviewed adequately.

\textbf{Confirming the analyses.} The signatures on analyses (FEA, thermal, electrical) confirm that the analyses were conducted, that the inputs were correct, that the results meet the requirements, that the limitations are understood. The signing engineer takes responsibility for the analyses' substance.

\textbf{Maintaining the design record.} The signed design and the supporting documents (analyses, calculations, test reports, decisions) are preserved as evidence of the signing's basis. The record supports subsequent review and the engineer's defence if the design's adequacy is later questioned.

\textbf{Refusing unsafe signatures.} The engineer who is asked to sign a design the engineer believes is inadequate is expected to refuse, to escalate the concern, and where necessary to use external channels (whistleblower protections; professional-body channels). The codes of ethics of the major engineering professional bodies (NSPE; IEEE; ASME; ASCE) include obligations of this character.

\textbf{Continuing competence.} The signing engineer maintains the competence the signature implies. Continuing professional education, engagement with the technical literature, and accumulation of practical experience are part of the obligation.

The Hyatt Regency walkway collapse (Volume X case) included engineering signatures on the design that was built; the post-failure analysis identified that the as-built design differed from the originally-analysed design in a way that the signing engineer should have caught. The professional sanctions following the collapse included loss of professional licences for the engineers involved. The case is a substantive example of the consequences of inadequate signature.

\begin{archetype}[ethics]
The engineering signature is the ethics archetype's working expression. The engineer's professional commitment to the design's adequacy is what the signature represents; the discipline of substantive signature is the discipline of making the commitment meaningful. Designs whose signatures are perfunctory commit the engineer's name without the underlying engineering work; the failures that follow reveal the gap between the form and the substance.
\end{archetype}

\section{Failure: rubber-stamp reviews and the legal aftermath}\pathtag{core}

The failure mode this section addresses is the design review that did not produce substantive engagement: the reviewers attended; the design team presented; questions were polite; concerns were not surfaced; the review passed. The design subsequently failed in service; the post-failure analysis identified issues the review should have caught.

The mechanisms of rubber-stamp reviews:

\textbf{Inadequate preparation.} The reviewers did not study the design materials in advance; arriving at the review, they could engage only at the most superficial level; substantive challenge requires depth the reviewers had not invested.

\textbf{Inappropriate reviewer selection.} The reviewers lacked the expertise to evaluate the design substantively; they could not identify the issues because they did not understand the design's domain or its specific characteristics.

\textbf{Cultural deference.} The reviewers deferred to the design team's expertise and did not challenge them; the cultural pattern of polite engineering meeting prevented substantive scrutiny.

\textbf{Schedule pressure.} The review was scheduled too tightly to allow substantive engagement; the time available was for ceremony rather than for substance.

\textbf{Pre-decided outcome.} The decision that the design would pass was implicit before the review began; the review was the formal validation of the pre-decided outcome. The reviewers' substantive engagement, even if it surfaced concerns, would not have changed the outcome.

\textbf{Diffusion of responsibility.} The review had many participants; each assumed others would raise concerns; no one did. The diffusion produces collective inaction even when individual participants would have raised concerns if responsible alone.

\textbf{Reviewer commercial interests.} The reviewers (typically other engineers in the same firm; sometimes external consultants paid by the design organisation) had commercial interests aligned with the design's passing; substantive challenge could threaten their relationship with the design organisation.

\textbf{Sponsor pressure.} A senior figure in the organisation has committed publicly to the design's success; the reviewers know that substantive challenge would create organisational difficulty; the review proceeds without the challenge.

The legal aftermath of failures following rubber-stamp reviews can be substantial:

\textbf{Professional liability.} The signing engineers face professional liability for the design's failure. The licensing body may discipline (suspension; revocation); the civil courts may award damages to harmed parties; in some cases, criminal charges may follow.

\textbf{Organisational liability.} The engineering firm faces corporate liability; the costs (damages; legal fees; reputation) may exceed insurance coverage; the firm's commercial survival may be threatened.

\textbf{Regulatory consequences.} The regulator may revoke certifications, impose corrective-action requirements, increase scrutiny on future submissions. The regulatory consequences extend beyond the specific design to the organisation's broader engagement with the regulator.

\textbf{Personal consequences.} Engineers personally responsible for design failures face career consequences (firing; difficulty re-employment); reputational consequences (the failure follows the engineer professionally); and in severe cases, criminal liability.

The Hyatt Regency walkway collapse engineers lost their professional licences and faced civil liability; the Therac-25 engineers' careers were affected; the Boeing 737 MAX engineers and the FAA's authorised representatives face ongoing legal and professional consequences. The legal aftermath is not deterrence-in-the-abstract; it is reality for engineers whose signatures are on failed designs.

The remediation against rubber-stamp reviews is the discipline of substantive review: explicit reviewer preparation requirements; appropriate reviewer selection; cultural commitment to substantive challenge; adequate time allocation; structural separation between the review's outcome and the project's commercial pressures; explicit individual responsibility for review findings; institutional protection of reviewers who raise concerns. The discipline costs effort and produces friction; the alternative is the failure mode this section addresses.

\begin{principle}[The signature commits the engineer]
The engineering signature is a professional commitment whose meaning is the engineer's responsibility for the design's adequacy. Reviews that do not produce substantive engagement, and signatures applied without substantive review, hollow out the commitment. The legal and professional consequences of failures following such patterns fall on the engineers whose names are on the documents; the discipline of refusing rubber-stamp roles is part of the engineer's professional self-protection.
\end{principle}

\begin{project}[Peer design review]
Pair with another reader of this volume; review each other's design from a previous chapter project (Chapter 5 detailed-design pack; Chapter 7 prototype iteration; Chapter 8 test plan). Conduct the review following the discipline of this chapter: prepare in advance by studying the materials; identify specific concerns; raise them in a structured session; document the concerns and the responses; track follow-up. Deliverable: the review document (concerns raised; responses received; resolutions) plus a one-page reflection on what the review surfaced that the design author had not anticipated.

\textbf{Hazard.} The exercise tests both the reviewer's discipline and the design author's openness to challenge. The point is the substantive engagement, not the polished outcome; reviews that surface significant concerns are more valuable than reviews that confirm the design without challenge.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A design review has 8 reviewers, each preparing for 4 hours and attending a 3-hour review session. The design team has 4 members each spending 2 hours in preparation and 3 hours in the session. Compute the total review effort in person-hours. Compare to the design's total development effort if the design took 2000 person-hours; identify the review fraction.
\end{exercise}

\begin{exercise}
An engineering firm has 50 engineers; 5 are licensed PE who can sign for the firm. If each licensed engineer can sign approximately 20 designs per year as the primary signing engineer, compute the firm's annual design output. Identify the implication for the firm's scaling.
\end{exercise}

\begin{exercise}
A certification submission requires 2000 pages of evidence; the regulator's review takes approximately 1 reviewer-day per 50 pages. Compute the regulator's review effort. If the regulator has 10 reviewers available, compute the calendar time for the review.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive a simple model of the relationship between review preparation effort and review effectiveness (the fraction of design issues surfaced). Identify the conditions under which additional preparation produces diminishing returns and the conditions under which it produces increasing returns.
\end{exercise}

\begin{exercise}
For a multi-reviewer review with each reviewer independently surfacing issues with probability $p$ per issue, derive the expected coverage as a function of reviewer count. Identify the conditions under which adding reviewers improves coverage substantially versus the conditions under which redundancy dominates.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the cost of a complete certification programme for a new commercial aircraft model. Identify the dominant cost categories (certification engineering; test evidence; regulatory fees; schedule impact).
\end{exercise}

\begin{exercise}
Estimate the cost of an independent review or IV\&V function on a major engineering programme. Compare to the cost of the design programme it reviews. Identify the conditions under which the IV\&V is justified.
\end{exercise}

\begin{exercise}
Estimate the fraction of major engineering failures (from Volume X cases or analogous) where the post-failure analysis identified inadequate design review as a contributing factor. Discuss what the fraction implies about the discipline as actually practised.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a PDR-CDR-FDR review protocol for an engineering organisation. Specify the participants, the materials required for each review, the criteria for passing, the documentation, and the integration with the project schedule.
\end{exercise}

\begin{exercise}
Design a substantive-review checklist for design reviewers. Include questions about: requirements coverage; analysis adequacy; failure modes; verification plan; manufacturing readiness; assumptions; risks.
\end{exercise}

\begin{exercise}
Design an independent review function for a hypothetical engineering organisation (e.g., a 200-engineer firm). Specify the function's reporting structure, its staffing, its role in design reviews, and the protections against the function being absorbed into the project line.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: the FAA's ODA arrangement for 737 MAX certification was structurally incapable of producing the substantive challenge the certification required \cite{acc:house-737max-2020}. Identify the structural reforms that would address the inadequacy and the political-economic constraints on implementing them.
\end{exercise}

\begin{exercise}
A senior engineer is asked to sign a design document for a design the engineer has not personally reviewed in detail (the engineer is relying on junior staff's review). Discuss the engineer's options under the professional-ethics frameworks. Identify the institutional structures that support or undermine the engineer's substantive engagement.
\end{exercise}

\begin{exercise}
Discuss the Hyatt Regency walkway case as a sign-off failure. Identify the specific failures of the design-review and sign-off processes and the institutional changes that would have caught the as-built-versus-as-designed discrepancy.
\end{exercise}

\begin{exercise}
Argue: the engineering profession's licensing structures (PE in US; CEng in UK; equivalents elsewhere) are necessary but insufficient protections against design failures. Identify what additional structures (institutional, cultural, regulatory) would substantively reduce the rate of failures attributable to inadequate sign-off.
\end{exercise}
