\chapter{Post-mortem as practice}
\label{vol10:ch16}

\epigraph{The post-mortem is the institution by which the engineering profession learns from itself.}{}

\chapmeta{Half-life: durable. Archetypes: ethics, failure. Exercise target: 10. Project track: hybrid.}

\noindent
This chapter is the volume's closing: the post-mortem as a working practice that aggregates the disciplines the volume has developed. The post-mortem is not the same as the formal accident investigation (Chapter 12), although it draws on the same evidence and reasoning. It is a smaller-scale, more frequent, more practitioner-driven activity that occurs after any project, incident, or near-miss; its purpose is to extract the lessons before they fade and to feed them into the next iteration. The chapter is about the practice as a craft, the institutional structures that support it, and the habit that distinguishes engineering organisations that learn from those that do not.

\section{The post-mortem as engineering practice}\pathtag{core}

The post-mortem is the structured retrospective conducted by the people who did the work, focused on what happened, why it happened, what worked, what did not, and what to do differently next time. The practice originated in military operations (where after-action reviews are mandatory after any significant engagement), in medical contexts (the morbidity and mortality conference; the surgical case review), and in spaceflight mission operations (every NASA mission has a post-flight review with the operating team). The software engineering profession has adopted the practice extensively in the era of pervasive incidents: Google's post-mortem culture, articulated in the Site Reliability Engineering book \cite{text:google-sre}, has been influential across the industry.

The structural elements of a working post-mortem practice:

\textbf{Triggered by event class, not by severity alone.} The post-mortem occurs after any event of a defined class: every customer-impacting production incident; every project that misses major milestones; every near-miss reaching a defined threshold; every decision that the team subsequently questions. The trigger is rule-based; the practice is not "we should probably post-mortem this," but "this event is in the class of events we post-mortem."

\textbf{Conducted by the people who did the work.} The post-mortem is not a top-down investigation; it is the working team's own retrospective. Outside facilitators may participate; the substantive content is the team's analysis of its own work.

\textbf{Time-bounded and document-producing.} The post-mortem is a structured meeting (typically 1-3 hours) that produces a written document. The document captures the timeline, the analysis, the lessons identified, and the actions committed. The document is the post-mortem; the meeting is the means of producing it.

\textbf{Action-oriented.} The post-mortem identifies specific actions to be taken, with owners and timelines. The actions are tracked; the post-mortem's value is contingent on the actions actually occurring.

\textbf{Aggregated over time.} Individual post-mortems are useful; aggregated patterns across many post-mortems are more useful. The aggregation surfaces recurring conditions that no single post-mortem would identify; the organisation learns from the pattern, not only from the individual events.

\textbf{Embedded in culture.} The post-mortem is not an exceptional event but a routine part of how the organisation operates. The cultural commitment is what distinguishes organisations that conduct post-mortems with substance from organisations that conduct them as ritual.

The relationship of the post-mortem to the formal accident investigation (Chapter 12) is the relationship of routine practice to exceptional practice. Both extract lessons from events; the post-mortem operates at higher frequency, lower formality, lower legal exposure, and within the team that did the work; the accident investigation operates at lower frequency, higher formality, often with legal consequences, and conducted by external investigators. The two practices are complementary: the post-mortem captures the lessons from the routine events; the accident investigation captures the lessons from the rare catastrophic events.

\section{Blameless post-mortems}\pathtag{core}

The blameless post-mortem is the explicit organisational commitment that the post-mortem will not be used to assign individual blame or to discipline individuals for the actions revealed in the analysis. The commitment is not a denial of accountability (individuals remain accountable for their decisions and actions through other organisational mechanisms); it is the specific separation of the post-mortem's learning function from the disciplinary function.

The reasoning for the commitment:

\textbf{Honest reporting requires safety.} A post-mortem in which participants fear personal consequences will not produce honest reporting; participants will minimise their own contributions, omit details that could be held against them, and shape the narrative to deflect responsibility. The blameless commitment is the structural condition for honest reporting.

\textbf{Individual blame is rarely the substantive cause.} As Chapter 9 developed, "operator error" attributions almost always rest on design, procedure, training, workload, or organisational conditions that produced the operator's action. Disciplining the individual without changing the conditions produces no improvement and removes the individual's contribution to future improvement.

\textbf{Cultural cost of blame culture is substantial.} An organisation that responds to incidents with individual blame produces a culture in which incidents are hidden, near-misses are not reported, post-mortems are perfunctory, and the systematic learning the practice could enable does not occur. The cultural cost is the systematic underperformance of the organisation's learning function.

The boundaries of the commitment are also explicit. The blameless commitment does not extend to: gross negligence (an individual deliberately ignored procedures with knowledge of consequence); intentional violation of safety rules; criminal activity; repeated patterns of error after correction. These cases are addressed through disciplinary processes that operate separately from the post-mortem; the post-mortem of an incident in which gross negligence occurred still proceeds blamelessly to extract lessons about the system, while the disciplinary process addresses the individual's accountability through other channels.

The institutional construction of a blameless culture is substantial work. The leadership commitment must be explicit and repeated; the practice must be modelled by senior practitioners conducting blameless post-mortems of their own decisions; the consequences of incidents must demonstrably not include individual punishment for the actions revealed in the analysis; the post-mortems must produce substantive change, demonstrating that the honest reporting is converted into improvement. The cultural state is not stable without active maintenance; it can be lost rapidly through inconsistent application or through senior decisions that contradict the commitment.

\begin{principle}[Blameless reporting, accountable systems]
The blameless post-mortem is the commitment that the post-mortem will produce learning, not punishment. Individuals remain accountable through other mechanisms; the post-mortem's accountability is to the system. The dual structure (blameless reporting; accountable systems) is what makes the learning function viable in the presence of the disciplinary function.
\end{principle}

\section{The five-whys and its critics}\pathtag{standard}

The "five whys" technique, originating in the Toyota Production System, asks "why?" iteratively after a problem is observed: why did the production line stop? Because the bolt broke. Why? Because it was over-torqued. Why? Because the torque gauge was miscalibrated. Why? Because the calibration interval had been extended. Why? Because the calibration-shop schedule had been compressed. The chain reaches a level at which the answer is a system condition that can be acted on, rather than a local proximate cause.

The technique's appeal is its simplicity and its tendency to drive the analysis beyond the proximate cause to the upstream conditions. Its appropriate use is in routine problem-solving where the causal chain is approximately linear and the actionable upstream condition is reachable in a few iterations.

The criticisms of the technique are substantive:

\textbf{Linear causal model insufficient for complex systems.} Many engineering failures arise from interactions of multiple conditions, not from a single causal chain. The five-whys technique, applied to such failures, produces a single causal chain that omits the other conditions. The Therac-25 incidents had a race condition, a hardware-interlock removal, an operator-interface design, and an organisational response to anomaly reports as contributing factors; a five-whys analysis of any single incident would produce a single chain through one of these and would miss the others.

\textbf{Confirmation bias and selection of the "first" why.} The technique provides no discipline for which "why" to pursue at each step; the choice is the analyst's, and the analyst's preferences (which conditions are easier to fix; which are politically acceptable to identify) bias the resulting chain. Different analysts produce different five-whys chains for the same incident; the technique does not converge.

\textbf{Stops at organisational boundary.} The technique tends to stop at the boundary of the analyst's organisational reach. A five-whys analysis conducted by an operations team stops at a development-team boundary; a five-whys analysis conducted by an internal team stops at the contractor-team boundary. The structural causes that cross organisational boundaries are systematically missed.

\textbf{Treats causes as discrete rather than as distributed conditions.} The five-whys technique presumes that there is a "cause" to be identified; the complex-systems framing (Snook; Hollnagel; Dekker) treats failures as the emergent property of distributed conditions that interact, with no single discrete cause. The technique's framing is inadequate to the complex-system reality.

The alternative methods address some of the criticisms. STAMP (Systems-Theoretic Accident Model and Processes), developed by Nancy Leveson \cite{text:leveson2011}, models accidents as the failure of safety constraints within a hierarchical control structure; the analysis identifies the constraints that were violated and the control loops that did not enforce them. The method is more capable than the five-whys for complex systems and is more transferable across organisational boundaries. AcciMap (Rasmussen and Svedung) models accidents as the failure of a hierarchical sociotechnical system, with explicit treatment of regulatory, organisational, and technical contributions; the method is well-suited to multi-stakeholder accidents.

The pragmatic position is that the five-whys technique is useful for routine problem-solving but inadequate for complex-system failure analysis; the analyst chooses the technique to match the problem complexity. The five-whys is a starting heuristic, not a complete methodology; its results are inputs to more substantive analysis when the problem warrants.

\section{STAMP and other systemic methods}\pathtag{standard}

STAMP (Systems-Theoretic Accident Model and Processes) is the systemic accident analysis method developed by Nancy Leveson and her collaborators at MIT \cite{text:leveson2011}. The framework treats the system being analysed as a hierarchical control structure: at each level, the level above imposes safety constraints on the level below; the level below provides feedback to the level above. Accidents occur when the safety constraints are violated; the analysis identifies the constraints that were violated and the control loops that did not enforce them.

The analysis steps:

\textbf{Identify the system's hierarchical control structure.} For an aviation system: ICAO at the top; national regulators (FAA, EASA, etc.); aircraft manufacturers; airlines; flight crews; the aircraft itself; the passengers. Each level controls the level below through specific mechanisms (regulations, contracts, procedures, displays, controls); each provides feedback to the level above through specific mechanisms (reports, inspections, surveillance, observation).

\textbf{Identify the safety constraints at each level.} What is each level required to ensure to maintain the system's safety? The flight crew must not allow the aircraft to enter an unrecoverable flight regime; the airline must operate within its certificate's limitations; the regulator must enforce the airworthiness standards; the manufacturer must build aircraft to certified specifications.

\textbf{Identify the violations of constraints that contributed to the accident.} For each level, what constraints did this level fail to enforce or violate?

\textbf{Identify the control-loop failures that produced the violations.} For each violation, what was missing or inadequate in the control loop? Inadequate feedback (the level above did not know the level below was violating)? Inadequate model (the level above had an inaccurate mental model of the level below)? Inadequate authority (the level above lacked the authority to compel correction)? Inadequate process (the level above had the authority and model but did not act)?

\textbf{Identify the recommended changes to the control structure.} What changes to the control loops, the constraints, the feedback mechanisms, the authority structure, would prevent the accident's recurrence?

The framework is more capable than linear cause-effect analysis for complex socio-technical systems. The 737 MAX accidents \cite{acc:house-737max-2020}, analysed via STAMP, surface the control-loop failures: Boeing's safety-assessment process did not adequately update when MCAS authority expanded (internal control loop); the FAA's ODA arrangement did not provide independent verification of the safety assessment (regulator-manufacturer control loop); the documentation to operators did not include MCAS (manufacturer-airline control loop); the training did not include MCAS failure recognition (airline-crew control loop). The accident occurred at the intersection of multiple control-loop failures; the recommended changes address each.

Other systemic methods (AcciMap; CAST, which is Leveson's STAMP-based analysis of single accidents; FRAM, Hollnagel's Functional Resonance Analysis Method) share the framing of accidents as system-level phenomena and the rejection of single-cause attribution. The methods are taught in safety-engineering programs and are increasingly used in major investigations; the practitioner community has built training and software-tool support for them.

\section{Writing a post-mortem document}\pathtag{standard}

A post-mortem document is itself an engineering artefact with developed conventions. The structure typically includes:

\textbf{Summary.} One paragraph: what happened, when, what was the impact, what is the highest-leverage action.

\textbf{Timeline.} The events in chronological order, with timestamps to the resolution the available evidence supports (often minutes for production-system incidents; sometimes seconds for safety-critical events; sometimes days for slow-moving incidents). The timeline is the factual basis from which subsequent analysis proceeds.

\textbf{Impact.} What was affected: users, systems, services, data, time. Quantitative where possible (number of affected requests; revenue impact; user-visible degradation duration; data-loss volume).

\textbf{Root-cause analysis.} The analytical section that identifies the contributing conditions. The depth depends on the incident's significance; routine incidents may have a few paragraphs; major incidents may have multi-page analyses. The analysis is structured (e.g., contributing factors enumerated; chain of conditions explicit); it is not narrative speculation.

\textbf{What went well.} The actions and conditions that limited the incident's impact or accelerated its resolution. Identifying what went well is as important as identifying what went poorly; the practices that worked deserve recognition and reinforcement.

\textbf{What went poorly.} The actions and conditions that worsened the incident or delayed its resolution. The discipline is to describe behaviours and conditions rather than to assign individual blame.

\textbf{Where we got lucky.} The conditions that prevented worse outcomes that could have occurred. "Lucky" outcomes are the conditions under which subsequent incidents may be less lucky; they warrant the same attention as the unlucky outcomes.

\textbf{Action items.} The specific actions committed, with owners and timelines. The actions are tracked separately (incident-tracking system; project-management tool); the post-mortem document is the snapshot at the time of writing.

\textbf{Lessons identified.} The general lessons that apply beyond the specific incident: design patterns that proved fragile; procedural assumptions that proved wrong; organisational conditions that contributed. These lessons feed into design standards, training, and process changes.

The post-mortem document is shared within the organisation (and sometimes externally, in companies that publish post-mortems as a transparency commitment and as a contribution to the broader engineering community). The publication discipline matters: a post-mortem written for internal audiences only is acceptable; a post-mortem written for external publication requires additional editorial review for confidentiality, legal, and competitive considerations.

\section{The post-mortem habit at organisational scale}\pathtag{standard}

The individual post-mortem is the unit of practice; the organisational habit is what makes the unit's accumulated effects substantial. The cultural and structural conditions that produce the habit:

\textbf{Time and attention.} Post-mortems take time: hours for the meeting, additional hours for writing the document, follow-up hours for the action items. The organisation must provide the time as a legitimate use of engineering effort; an organisation in which engineers are too busy to post-mortem is an organisation that does not have time to learn.

\textbf{Tracked action items.} The actions committed in post-mortems are tracked, with regular review of completion. Actions that are not completed (because the assignee left, because priorities shifted, because the action proved harder than anticipated) are explicitly addressed; the alternatives are quiet abandonment, which corrodes the practice.

\textbf{Aggregated review.} Periodic review of recent post-mortems identifies patterns: recurring contributing factors; recurring action-item categories; recurring conditions that the previous post-mortems' actions did not address. The aggregated review is the organisation's higher-level learning loop; it produces the standards changes, training changes, process changes that no individual post-mortem would surface.

\textbf{Public access and reading.} Engineers in the organisation read each other's post-mortems; the practice spreads the lessons across teams; new engineers absorb the patterns by reading the corpus. Organisations with a strong post-mortem culture often maintain searchable databases of post-mortems that new engineers are expected to read substantial subsets of during onboarding.

\textbf{Leadership participation.} Senior engineers and leaders participate in post-mortems, both of their own decisions and as facilitators of others'. The participation models the practice and demonstrates the leadership commitment; absent participation signals that the practice is for lower-status engineers, which corrodes the cultural commitment.

\textbf{Continuity through personnel changes.} The post-mortem corpus and the action-item tracking persist across personnel changes; departing engineers transfer their open action items to other owners; the organisational memory is in the documents and tracking, not only in the individuals who participated.

The organisations that have built strong post-mortem cultures (Google in the SRE tradition \cite{text:google-sre}; Etsy and the broader DevOps community; the major cloud providers; some financial-services firms; many nuclear operators in the post-TMI tradition; the US military's after-action review tradition) demonstrate that the practice can be sustained at scale and that its accumulated effects are substantial. The cumulative learning from years of post-mortems shapes the organisation's reliability, its engineering norms, its ability to operate complex systems with adequate safety.

\section{Closing: handing the reader to Volume XI}\pathtag{core}

This volume has been the volume of failure. The chapters have developed the engineering taxonomy of failure (Chapter 1), the physical mechanisms (Chapters 2-4), the software and concurrency mechanisms (Chapters 5-6), the network and systemic mechanisms (Chapters 7-8), the human and organisational dimensions (Chapters 9-10), the regulatory and forensic disciplines (Chapters 11-12), and the practices that learn from failure prospectively and retrospectively (Chapters 13-14 and 16), with the planetary-scale failures (Chapter 15) as the contemporary frontier.

The volume's thesis is that engineering is the discipline by which measured reality becomes reliable intervention under constraint, failure, scale, and responsibility, and that failure is the first-class subject from which the engineering profession constructs its discipline. The successful engineer is not the engineer who avoids failure (impossible) but the engineer who builds systems whose failures are bounded, whose modes are understood, whose lessons are extracted, and whose recurrences are progressively reduced. The chapters of this volume aggregate to the working tools for that engineer.

Volume XI takes the engineering practice into the procurement, contracts, project-management, and economic context that determines what gets built, at what cost, on what schedule, for whom, and with what consequences. The transition is from the technical and human-organisational level (this volume's scope) to the financial and contractual level (Volume XI's scope). The engineer who has internalised the discipline of failure in this volume's sense is the engineer Volume XI prepares to operate in the world of clients, budgets, and contracts. The arc continues; this volume is its centre.

The reader who has worked through this volume should be able to: identify the failure modes plausible for a system being designed; assess the failure-handling adequacy of a system in operation; conduct a competent post-mortem of an incident; read an accident-investigation report critically; identify the organisational and regulatory conditions that produce or prevent failures; participate substantively in the engineering profession's responsibility for failure across all of its scales.

The work is engineering. The work is also the engineering profession's working ethic. The two are the same.

\begin{principle}[The discipline of failure is the discipline of engineering]
Engineering's specific contribution to human knowledge and practice is the discipline of building systems that work under constraint, fail in bounded ways, and improve over generations of accumulated learning from failure. The discipline of failure is not a sub-discipline of engineering; it is the discipline that distinguishes engineering from other forms of building.
\end{principle}

\begin{archetype}[failure]
The post-mortem is the archetype's reflexive expression: the discipline of failure applied to the engineering practice itself. The engineer who post-mortems the engineer's own work is the engineer whose practice is engineering rather than mere production.
\end{archetype}

\begin{project}[Post-mortem of one project from this book]
Select one project from the projects assigned across this book's volumes that the reader has attempted and that did not work as the reader expected (a project whose outcome was different from what was planned; a project whose execution revealed assumptions that proved wrong; a project that was abandoned before completion). Write a post-mortem of the project following the structure of this chapter. Include: the timeline of the work; the expected versus actual outcome; the contributing conditions to the divergence; what went well; what went poorly; what the reader would do differently next time; what the reader has learned that applies beyond this project. Deliverable: a three-page post-mortem document, written to the discipline this chapter has developed.

\textbf{Hazard.} The blameless commitment applies to the reader's own work. The point of the exercise is not self-criticism but learning; the document is for the reader's own future practice and for whatever colleagues the reader chooses to share it with. The professional register matters even when the audience is the reader alone.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A team conducts 50 post-mortems per year; each takes 4 hours of meeting time plus 3 hours of writing time, with 5 participants per meeting and 1 author per document. Compute the annual engineering-hour investment in the post-mortem practice. Compare to the investment in initial design and code review for an equivalent volume of work.
\end{exercise}

\begin{exercise}
An organisation's post-mortem-corpus database grows by 50 incidents per year, with each post-mortem averaging 5 pages. Compute the corpus size after 5 years. Discuss the implications for search, retrieval, and aggregated review.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
For a STAMP-style control-loop analysis, derive the conditions under which a control loop fails: the controller does not receive feedback; the controller's model of the controlled process is inaccurate; the controller's control actions are inadequate; the controlled process behaves outside its constrained envelope. Identify the design changes that close each gap.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the productivity benefit of a strong post-mortem culture for an engineering organisation of 100 engineers operating complex systems. Discuss the dependence on the rate of incidents that the culture would otherwise produce and on the half-life of the lessons learned.
\end{exercise}

\begin{exercise}
Estimate the time required for an engineer to develop a working post-mortem habit (the practice of conducting blameless retrospectives on their own work after each significant project). Identify the institutional supports that accelerate the development.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a post-mortem template for an engineering team. Specify the sections, the required content per section, the metadata captured for aggregated review, and the publication policy.
\end{exercise}

\begin{exercise}
Design an aggregated-review process for an organisation's post-mortem corpus. Specify the review cadence, the analytical methods, the deliverable, and the integration with the organisation's standards, training, and process changes.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: the blameless post-mortem commitment is incompatible with individual accountability and should be replaced with a more nuanced framework. Argue the opposite. Identify the conditions under which each position is the right one to take.
\end{exercise}

\begin{exercise}
Discuss the limitations of the five-whys technique for complex socio-technical failures. Identify the conditions under which the technique is appropriate, the conditions under which it is inadequate, and the alternatives that should be used in the inadequate cases.
\end{exercise}

\begin{exercise}
The discipline of post-mortem is, in this chapter's framing, the discipline of failure applied reflexively to the engineering practice itself. Argue that the discipline distinguishes engineering as a profession from production as an activity; identify the practical implications for engineering education, professional registration, and institutional accountability.
\end{exercise}
