\chapter{The future of the artificial}
\label{vol12:ch14}

\epigraph{The artificial world is the world we have made; the future of the artificial is the world we will make next.}{Author}

\chapmeta{Half-life: medium-life. Archetypes: all, ethics, scaling. Exercise target: 0. Project track: written (10-page personal manifesto on the engineering work the reader intends to do, defending it with everything in the previous $\sim 18{,}000$ pages). Hazard class: medium (forward-looking; openly normative). Reader path default: all sections core. Named cases: the volume's preceding cases are recalled and revisited; no new named cases are introduced.}

\noindent
The book ends here. The chapter is the closing of the twelve volumes. It is openly forward-looking and openly normative; it returns to the project's thesis and to the reader's relation to it. The chapter has no exercises and a single project: a personal manifesto in which the reader articulates the engineering work they intend to do and defends it with the resources of the preceding $\sim 18{,}000$ pages. The manifesto is the closing exercise of the book and is the test of whether the project has accomplished what it set out to accomplish.

\section{The artificial world as it stands}\pathtag{core}

The artificial world as it stands in 2026 is the cumulative engineering output of $\sim 250$ years of industrial activity and several thousand years of pre-industrial engineering. It is approximately $1$ trillion tonnes of engineered materials and structures, distributed across approximately $200$ political jurisdictions on a planet of $\sim 8$ billion humans. It is approximately $30$ million km of paved road, $1.5$ million km of submarine cable, $10$ million km of high-voltage transmission line, $50{,}000$ large dams, $25$ million km of municipal sewer, $440$ operating nuclear reactors, $10{,}000$ active satellites, an internet of $\sim 5$ billion devices, a healthcare system of $\sim 16$ million physicians and $\sim 30$ million nurses, an aviation system of $\sim 25{,}000$ commercial aircraft serving $\sim 4$ billion passengers per year, a manufacturing capacity producing $\sim 90$ million cars per year, a food system feeding $\sim 8$ billion humans, an information system processing $\sim 5$ zettabytes per year. The artificial world is, by the measures of the previous twelve volumes, the largest single intervention humans have made in the bio-physical environment.

It is also, by the measures of the same volumes, an intervention whose long-term sustainability is in genuine question. The climate boundary is crossed; the nitrogen and phosphorus boundaries are crossed; the biosphere-integrity boundary is crossed; the freshwater-and-land-use boundaries are crossed; the novel-entities boundary is crossed. The contemporary engineering response is at a particular moment in the multi-decade trajectory: the technical pathways for substantial response are established; the political-economic deployment is uneven; the cumulative outcome over the next $50$-$100$ years remains uncertain. The reader who has finished the previous twelve volumes has the engineering literacy and the civilizational understanding to engage with the trajectory as a substantive participant rather than as an observer.

\section{The trajectories: techno-optimist, techno-pessimist, techno-realist}\pathtag{core}

The contemporary public-and-engineering discourse identifies several trajectories for the artificial world over the next $50$-$100$ years. The techno-optimist trajectory: continued technological progress; the energy transition is completed; AI-and-automation produces substantial productivity gains; the developing world reaches developed-world income levels; the planetary boundaries are managed through engineered responses; substantial human flourishing in the second half of the 21st century. The techno-pessimist trajectory: the energy transition is too slow; climate damage is substantial; AI-and-automation displaces labour faster than alternatives emerge; the developing world's energy-and-food access remains incomplete; the planetary boundaries push beyond recovery in several dimensions; substantial human suffering in the second half of the 21st century. The techno-realist trajectory: a mixed outcome with substantial progress in some dimensions and substantial failure in others; the engineering profession's role is to maximise the progress and minimise the failure across the trajectory.

The project's posture is techno-realist with explicit acknowledgement of the contingency. The contingency is in the political-economic execution: the engineering profession can document what is technically achievable; the deployment depends on political-economic decisions the profession does not control. The project's argument throughout has been for the engineering profession to engage with the political-economic context, both because the engineering work depends on it and because the engineering profession's contribution to the political-economic context is one of the determinants of the eventual outcome.

The reader of this chapter is asked to form their own view of the trajectories and the contingency. The chapter does not insist on the techno-realist view; it argues for it and provides the resources of the previous twelve volumes as the basis for evaluation. The reader who arrives at a different view has used the resources to do so; that is the project's deeper aim than the assent to any particular position.

\section{What only the next generation will see}\pathtag{core}

Several elements of the artificial world are most relevant for the next generation rather than the current one. The energy transition's completion (assuming it proceeds on the contemporary trajectories) will be in the 2050s-2060s, with the affected populations being substantially the next generation. The full climate consequences of the cumulative emissions through the 2030s will manifest in the 2060s-2100s, with the affected populations being primarily the next generation and the ones after. The biosphere-integrity trajectory will be determined by decisions made now whose consequences will be lived by the next generation. The space-infrastructure expansion (LEO constellations; lunar return; Mars considerations) will reach maturity in the 2050s-2070s, with the affected populations being the next generation. The AI trajectory's substantial consequences (whatever they are) will be substantially in the next generation's lifetime.

The contemporary engineering profession's relation to the next generation: the engineering decisions made now substantially determine the conditions the next generation will inherit. The discipline this implies: explicit long-horizon thinking that takes the next-generation perspective seriously rather than the political-cycle perspective that the contemporary institutional structures over-weight. The engineering profession has institutional structures (professional societies; academic institutions; long-horizon research programmes) that can support the long-horizon thinking; the project's argument is that the profession should use them.

\section{What the engineer's child should learn}\pathtag{core}

The engineering profession's transmission to the next generation is partly through formal education and partly through the example the contemporary profession sets. The formal-education question is the substance of the engineering-education debates that the project has touched on throughout: the integration of the technical and the civilizational; the integration of the failure literacy with the design discipline; the explicit attention to the political-economic context; the explicit attention to the planetary-boundary context. The contemporary engineering-education trajectory in most jurisdictions is incrementally moving in this direction with substantial variation in pace and depth.

The example-the-contemporary-profession-sets question is the substance of the chapter's normative position. The engineer's child learns engineering not principally from the engineering curriculum but from the engineering culture: how engineers talk about their work; what engineers admit to having got wrong; how engineers engage with the consequences of their work; whom engineers blame and whom engineers credit; what engineers consider their professional responsibility and what they consider out of scope. The engineer's child who sees the contemporary profession engaging credibly with the civilization-scale responsibilities the project has documented will form one understanding of what engineering is; the engineer's child who sees the contemporary profession avoiding the engagement will form a different one. The chapter's argument: the profession's collective example is one of the determinants of the next generation's engineering profession, and the contemporary engineers bear responsibility for that example.

\section{The engineer in the world to come}\pathtag{core}

The engineer in the world to come will inherit a context in which the climate-system disruption is more advanced than today's; the planetary-boundary excursions are more consequential; the energy-transition is partly completed; the AI-and-automation deployment is more pervasive; the population is larger and more urbanised; the geopolitical context is different in ways the contemporary view cannot fully anticipate. The engineering work the next generation does will be partly continuous with the present generation's (the build-and-maintain of the infrastructure the previous volumes have described) and partly discontinuous (the response to the climate-and-boundary consequences; the AI-and-automation engineering scope; the unforeseen developments).

The engineering profession's institutional and cultural inheritance to the next generation: the technical literacy; the failure literacy; the design discipline; the civilization-scale awareness. The next generation will adapt the inheritance to the context they encounter; the contemporary generation's contribution is the inheritance itself. The project's hope: the inheritance is adequate to the challenges the next generation will face.

\section{Closing the twelve volumes: the thesis returns}\pathtag{core}

The book's thesis, stated at the top of \texttt{docs/editorial-decisions.md} and pinned throughout: ``Engineering is the discipline by which measured reality becomes reliable intervention under constraint, failure, scale, and responsibility. Engineering is not applied science. Engineering is disciplined intervention.''

The twelve volumes have been the elaboration of the thesis across the engineering disciplines and contexts. Volume I established the mathematical foundation of measured reality. Volumes II-V developed the materials, forces, energy, and dimensions of physical reality the engineering profession works with. Volume VI extended to biology and the living systems. Volume VII added the information dimension. Volume VIII covered the machines and automation. Volume IX integrated the systems. Volume X documented the failure literacy that is the honest engineering profession's discipline. Volume XI codified the design discipline through which engineering work is actually done. Volume XII has placed all of the preceding in the civilization-scale context.

The thesis returns in this closing chapter as the integration of the volumes' contributions. ``Measured reality'': the technical literacy of Volumes I-IX. ``Reliable intervention'': the design discipline of Volume XI plus the failure literacy of Volume X. ``Under constraint, failure, scale, and responsibility'': the civilization-scale awareness of Volume XII. ``Engineering is disciplined intervention'': the project's summary of what engineering is, the alternative to the ``applied science'' framing that the project has rejected throughout.

\section{The author's posture: what the reader keeps after the book}\pathtag{core}

The author's posture across the project: the engineering profession is one of the central professions of civilization and bears continuing responsibility for civilization's bio-physical and infrastructural condition. The profession's traditional self-understanding is incomplete for the contemporary moment, which requires the integration of the technical and the civilizational. The integration is achievable through the discipline the project has tried to teach: the technical literacy, the failure literacy, the design discipline, the civilizational awareness.

The reader keeps after the book: the technical literacy of an engineer (the mathematical, physical, biological, informational, mechanical, systems-engineering foundation); the failure literacy of an honest one (the recognition that engineering work fails in characteristic ways and the discipline to design against the failure modes); the design discipline of a capable one (the actual process by which engineering work is conducted); the civilizational understanding that distinguishes engineering-as-profession from engineering-as-technical-activity-for-hire. The reader who has finished the book has, by the project's measures, the foundation for the engineering profession's contemporary substantive role.

The reader is also asked to engage substantively with the parts of the project they disagree with. The project has been openly opinionated where the editorial decisions demanded; the reader is invited to identify the positions they reject and to articulate why. The project's deepest hope is for the reader to engage with the engineering profession's substance rather than to accept the project's positions without engagement.

\section{The book ends; the work does not}\pathtag{core}

The book ends here. The reader who has finished the twelve volumes has read approximately $18{,}000$ pages across two-and-a-half years of dedicated engagement (per the project's assumed reading pace) or longer at a less dedicated pace. The reader has done the work of acquiring the engineering literacy the project set out to teach.

The work of being an engineer, however, does not end with the book. It begins with the book. The engineering profession is a multi-decade vocation in which the literacy the book has provided is the starting condition rather than the completion condition. The reader who has finished the book is asked to take the literacy into the engineering work they do, into the engineering profession's institutional engagement, into the political-economic context in which the engineering work is conducted, into the civilizational responsibility the engineering profession bears.

The book's last sentence: do the work. The civilization that engineered itself into the conditions documented in this volume can engineer itself into different conditions. The engineering profession that produced the failures documented in Volume X and the design discipline codified in Volume XI is the profession whose continuing work over the next decades will determine the trajectory. The reader is part of that profession. The work is yours.

\section{Project}\pathtag{core}

\begin{project}[Personal engineering manifesto]
Write a ten-page personal manifesto on the engineering work you intend to do. Defend it with everything in the previous $\sim 18{,}000$ pages of this book. Specifically: identify the engineering domain and the specific work scope you intend to pursue; articulate the civilizational contribution the work makes; identify the failure modes the work must protect against and the design discipline you will bring to it; identify the political-economic context in which the work will be conducted and your intended engagement with it; identify the planetary-boundary and just-transition considerations and your intended response; identify the engineering profession's institutional structures with which you intend to engage; identify the limits of what your individual work can accomplish and the collaborations and institutional changes you intend to pursue. Defend each element with reference to the engineering substance of the preceding volumes; identify the project's positions you accept, the positions you reject, the positions you modify. The manifesto's purpose is the integration of the engineering literacy the book has tried to teach with the engineer's own substantive engagement; the manifesto is the test of whether the project has accomplished what it set out to accomplish, and it is also the start of the engineering work the reader will do.
\end{project}
