\chapter{Error and uncertainty}
\label{vol01:ch04}

\epigraph{The number is half the measurement; the uncertainty is the
other half. A reported value without a stated uncertainty is a
fragment.}{The working metrologist's rule, stated in the vocabulary
of the GUM \cite{std:iso-iec-guide98} and the VIM
\cite{std:jcgm-vim}.}

\chapmeta{Half-life: foundational. Archetypes: uncertainty (first
formal introduction in Volume I); balance (dimensional balance
recurs in propagation derivations). Exercise target: 35. Project
track: household + analysis (hybrid). Hazard class: low. Reader
path default: standard, with sections explicitly tagged. Named
cases: Hubble Space Telescope primary mirror (1990), revisited from
chapters 2 and 3 as an auxiliary-test override case in the failure
section; Vancouver Stock Exchange index rounding (1982-1983) in the
significant-figures section as the accumulation-of-truncation
illustration.}

\noindent
A measurement returns a value and an uncertainty. The value says
where the measurement lands; the uncertainty says how wide the
defensible interval is. This chapter makes the second quantity
operational: we compute it, propagate it through measurement models,
and reject uncertainty figures selected for institutional convenience
rather than derived from evidence.

\begin{mastery}
\textit{Historical sketch: how the GUM came to be.}
Before the late 1970s, working metrology laboratories around the
world used different formal conventions to report uncertainty. A
calibration certificate issued in West Germany used DIN 1319's
"Fehler" classification; a certificate from the National Bureau of
Standards in the United States used a "random plus systematic"
budget with different propagation rules; the United Kingdom's
National Physical Laboratory used its own conventions. The
disagreements meant that a downstream user comparing two
calibrations had no shared vocabulary for the answer.

The Bureau international des poids et mesures (BIPM) convened a
working group in 1977 (CIPM Recommendation INC-1) to harmonise
the conventions. The recommendation introduced the Type A
(statistical) versus Type B (other-information) classification
the chapter uses below, and asserted that all uncertainty
components be combined by quadrature into a single combined
standard uncertainty. The recommendation was non-binding but was
adopted by the ISO Technical Advisory Group, which produced the
first edition of the \textit{Guide to the Expression of
Uncertainty in Measurement} (GUM) in 1993. The guide has been
revised repeatedly (the current form is the 2008 corrected
reissue, JCGM 100:2008) and is the international reference
\cite{std:iso-iec-guide98}. A complementary document, the
International Vocabulary of Metrology (VIM, JCGM 200), gives the
standardised terminology \cite{std:jcgm-vim}.

Two follow-on supplements matter for working practice. JCGM 101
(2008) describes Monte Carlo propagation of distributions as the
alternative when the GUM's linear approximation fails
\cite{std:jcgm-101}; this chapter's section 4.3 uses it. JCGM 102
(2011) extends both forms to multivariate output quantities. The
chapter quotes the GUM's algebraic rules in working form; a
serious calibration report cites the GUM by edition and uses its
language verbatim.
\end{mastery}

\section{Random vs systematic error}\pathtag{core}

The international vocabulary of metrology distinguishes
\engterm{random error} (a component of measurement error that
varies in an unpredictable manner across repeated measurements) from
\engterm{systematic error} (a component of measurement error that,
in repeated measurements, remains constant or varies in a
predictable manner) \cite{std:jcgm-vim}. The distinction is
operational: it determines which mitigation works.

Independent random error is reduced by averaging. If a quantity is
measured $n$ times under repeatability conditions and the random
component is independent across measurements, the standard
uncertainty of the mean is the sample standard deviation divided by
$\sqrt{n}$. The central-limit-theorem argument is taken up below; the
operational consequence is narrower than the slogan: more data
reduces only the independent random component.

Systematic error is not reduced by averaging. If a thermometer reads
$0.2 \,\si{\degreeCelsius}$ high because of a calibration offset,
averaging a thousand readings produces a mean that is still
$0.2 \,\si{\degreeCelsius}$ high. The mechanism that reduces
systematic error is calibration against a reference that does not
share the offset, plus correction or adjustment. The discipline
that prevents systematic error from being mistaken for random error
is the separation of the two in the uncertainty budget.

\input{volumes/01-quantity/ch04-error-and-uncertainty/figures/fig-random-vs-systematic.tex}

\Cref{fig:vol01:ch04:random-vs-systematic} contrasts the two
patterns. The discipline of separating them in the budget pays off
because the mitigation strategies are different and the
arithmetical limits on reducing each are different.

\subsection{Mechanisms}

The mechanism matters because it tells us which corrective action has
a chance of working.

Random-error mechanisms include thermal noise in electronic
instruments, quantisation noise in digital instruments (the last
digit fluctuates because the underlying quantity is rarely exactly
on a quantisation step), parallax in analog instruments,
hand-tremor in manually-positioned instruments, repeatability
limits in mechanical instruments, and the effects of brief
environmental fluctuations.

Systematic-error mechanisms include calibration offset, calibration
drift between recalibrations, instrument-specific bias from
component aging, environmental gradient (a temperature gradient
across a measurement system biases all readings in the same
direction), method-specific bias (the chosen procedure systematically
under- or overestimates the quantity), and the operator's
interpretive bias.

A mechanism that begins as random can become systematic if the
underlying cause shifts. A thermometer in a draft has random
fluctuations from the airflow; a thermometer in a steady-state warm
room has a systematic bias from the room's temperature gradient.
The distinction is in the time scale and the consistency of the
underlying cause, not in the mechanism's name.

\subsection{Type A and Type B in this language}

Chapter 1 introduced the GUM's Type A and Type B distinction at a
teaching level \cite{std:iso-iec-guide98}. We restate it here in
the language of random and systematic error.

\engterm{Type A evaluation} estimates uncertainty by statistical
analysis of observed data, usually repeated measurements. If the
independent random component has standard deviation $\sigma$ in a
single measurement, the standard uncertainty of the mean of $n$
measurements is estimated as $s/\sqrt{n}$.

\engterm{Type B evaluation} estimates uncertainty from other
information: instrument specifications, calibration certificates,
manufacturer data, documented experience, previous measurements, and
physical models. A calibration correction taken from a certificate
is commonly a Type B contribution, even when the effect being
corrected is a systematic bias.

The distinction between Type A and Type B is the method of
evaluation, not the random-versus-systematic character of the
effect. A bias can be estimated by Type A methods when repeated
calibration data, intercomparisons, or controlled condition changes
expose it statistically. A random contribution can enter a Type B
evaluation when previous data, a specification, or a physical model
gives the standard uncertainty without a new repeated-measurement
campaign.

\begin{archetype}[Uncertainty, first formal introduction]
Uncertainty is the engineering habit of carrying a second quantity
beside the reported value. The value says where the measurement
lands; the uncertainty says how wide the defensible landing zone is.
Later chapters attach the same habit to structural loads,
temperatures, material properties, reliability estimates, and
failure rates. In this chapter the pattern becomes formal: a number
reported without its uncertainty is not yet a usable measurement.
\end{archetype}

\section{Distribution of error: Gaussian, uniform, asymmetric}\pathtag{core}

A standard uncertainty is a single number that summarises a
distribution. The distribution itself carries more information.
Three distribution shapes recur in engineering measurement.

\subsection{Gaussian}

The Gaussian (normal) distribution, $\mathcal{N}(\mu, \sigma^{2})$,
is the canonical distribution of measurement error in the limit of
many small independent contributions. Its probability density is

\[
p(x) = \frac{1}{\sigma\sqrt{2\pi}} \exp\!\left(-\frac{(x - \mu)^{2}}
{2\sigma^{2}}\right).
\]

The Gaussian is fully characterised by its mean $\mu$ and its
standard deviation $\sigma$. The probability that a single draw
falls within $\pm 1\sigma$ of the mean is approximately $0.683$;
within $\pm 2\sigma$, $0.954$; within $\pm 3\sigma$, $0.997$. These
coverage probabilities map directly to the reporting conventions
introduced in section 4.6.

The Gaussian is appropriate when the underlying error mechanism is
the sum of many small independent contributions, none of which
dominates. The central limit theorem (section 4.4) guarantees the
limiting distribution under those conditions.

\subsection{Uniform}

The uniform distribution on an interval $[a, b]$ has density

\[
p(x) = \frac{1}{b - a} \quad \text{for } a \leq x \leq b, \quad 0
\text{ otherwise}.
\]

Its mean is $(a + b)/2$ and its standard deviation is
$(b - a)/(2\sqrt{3})$. The uniform distribution is appropriate when
the error is bounded above and below and the available information
gives no preference for any value within the bounds. Quantisation
error is often modelled as uniform over one quantisation interval
when the input is not locked to the quantiser and sufficient
variation or dithering is present. A specification that gives a
maximum and minimum without a most-likely value is commonly treated,
under the GUM's rectangular-distribution convention, as uniform
between those bounds \cite{std:iso-iec-guide98}.

A frequent Type B evaluation: a manufacturer states a maximum error
of $\pm a$ without a distribution. The standard uncertainty is then
$a/\sqrt{3}$, computed from the uniform distribution on
$[-a, +a]$.

\subsection{Asymmetric distributions}

The following examples are standard statistical models, not automatic
assignments. The measurement report should name the mechanism or
cite the convention that justifies the distribution
\cite{text:taylor1997}.

\begin{itemize}[leftmargin=*]
  \item \engterm{Lognormal}: the logarithm of the quantity is
    Gaussian. Multiplicative-error mechanisms (compound effects,
    biological growth, financial returns, particle-size
    distributions) often produce lognormal errors.
  \item \engterm{Half-normal}: the absolute value of a Gaussian.
    Errors that are bounded below by zero (deviations, magnitudes,
    distances to a reference) sometimes follow this distribution.
  \item \engterm{Weibull}: lifetime distributions, fatigue-cycle
    counts, time-to-failure data.
  \item \engterm{Poisson}: counting processes at low rates (radiation
    counts, defect counts, network packet arrivals). For a Poisson
    distribution with mean $\lambda$, the variance is also
    $\lambda$ and the standard deviation is $\sqrt{\lambda}$.
\end{itemize}

For asymmetric distributions, reporting a single $\pm u$ figure
loses information. The discipline is to report either both bounds
explicitly or the median plus the asymmetric percentile range.

\subsection{Heavy tails}

Some distributions have tails heavy enough that the ordinary
finite-variance central limit theorem does not apply. The Cauchy
distribution is the clean example: it has no finite mean and no
finite variance, and the average of independent Cauchy variables
remains Cauchy-distributed. A measurement model with Cauchy-like
error is not repaired by taking more ordinary averages. The
practical rule is simple: before applying Gaussian tools, check that
the variance is finite and that the observed tails are compatible
with the model.

\section{Propagation of uncertainty: linear and Monte Carlo}\pathtag{core}

Most engineering quantities are computed from measured inputs. A
volume is computed from a measured radius and a measured height; an
acceleration is computed from a measured velocity change and a
measured time interval; a power is computed from a measured voltage
and a measured current. Each input has its own uncertainty. The
output's uncertainty must be computed from the inputs'
uncertainties. The discipline that does the computation is
\engterm{propagation of uncertainty}.

\subsection{Algebraic propagation rules}

A reader at this point in Volume I has algebra and trigonometry but
not yet calculus, which Volume II takes up from first principles.
The algebraic propagation rules below cover the bulk of practical
uncertainty propagation a working engineer encounters; the
generalisation to arbitrary functions of several inputs uses partial
derivatives and is treated formally in Volume II Chapter 4. The
GUM \cite{std:iso-iec-guide98} states the general formula for
reference, included as a mastery box at the end of this section; for
calculations in this chapter we use the algebraic special cases.

\engterm{Sum and difference}. For $y = a x_{1} \pm b x_{2}$ with
$a, b$ constants and uncorrelated inputs $x_{1}, x_{2}$,

\[
u_{c}^{2}(y) = a^{2} u^{2}(x_{1}) + b^{2} u^{2}(x_{2}).
\]

The uncertainties of additive terms combine in quadrature, weighted
by the squared coefficients. A useful consequence: if two nominally
equal-uncertainty terms add, the combined standard uncertainty is
$\sqrt{2}$ times either term, not twice it.

\engterm{Product and ratio}. For a product of powers
$y = a x_{1}^{p_{1}} x_{2}^{p_{2}} \cdots$ with uncorrelated inputs
and small relative uncertainties, the relative uncertainty satisfies

\[
\left(\frac{u_{c}(y)}{y}\right)^{2} = \sum_{i} p_{i}^{2}
\left(\frac{u(x_{i})}{x_{i}}\right)^{2}.
\]

The exponent $p_{i}$ on each input is the input's
\engterm{sensitivity coefficient} in the relative-uncertainty rule:
the contribution to the combined relative uncertainty is the
exponent squared, times the relative uncertainty of the input,
squared, summed. For a simple product $y = x_{1} x_{2}$ both
exponents are $1$, so the relative uncertainties add in quadrature.
For $y = x^{2}$, the exponent is $2$ and the relative uncertainty
doubles before quadrature. For $y = \sqrt{x}$, the exponent is
$1/2$ and the relative uncertainty halves. For $y = 1/x$, the
exponent is $-1$ and the relative uncertainty enters with the same
magnitude (the sign drops in the square).

If the inputs are correlated, covariance terms must be added; the
simple quadrature rule no longer applies.

\engterm{Correlated difference: common-mode cancellation}. For
$y = x_{1} - x_{2}$ with correlation coefficient $\rho$,

\[
u_{c}^{2}(y) = u^{2}(x_{1}) + u^{2}(x_{2}) - 2\rho u(x_{1}) u(x_{2}).
\]

When $\rho = +1$ and $u(x_{1}) = u(x_{2})$, the common uncertainty
component cancels. Difference measurements exploit this: subtracting
two readings from the same drifting instrument can reject the shared
drift while preserving the change between the readings. Comparator
measurements in metrology are built around this common-mode
cancellation.

\begin{mastery}
The general GUM linear propagation formula. For an output quantity
$y$ that is an arbitrary function $y = f(x_{1}, x_{2}, \ldots,
x_{N})$ of $N$ input quantities $x_{i}$ with standard uncertainties
$u(x_{i})$, the combined standard uncertainty satisfies

\[
u_{c}^{2}(y) = \sum_{i=1}^{N}\left(\frac{\partial f}{\partial x_{i}}
\right)^{2} u^{2}(x_{i}) + 2\sum_{i=1}^{N-1}\sum_{j=i+1}^{N}
\frac{\partial f}{\partial x_{i}} \frac{\partial f}{\partial x_{j}}
u(x_{i}, x_{j}),
\]

where $u(x_{i}, x_{j})$ is the covariance between inputs $i$ and
$j$, and the partial derivatives are evaluated at the input
estimates. The formula is a first-order Taylor approximation. It
specialises to the algebraic rules above for sums (linear $f$) and
products (logarithmic-derivative manipulation). Volume II Chapter 4
introduces partial derivatives and Taylor expansion; the general
formula and its non-algebraic applications belong there. For
calculations in this volume, the algebraic rules above suffice.
\end{mastery}

\begin{estimation}
Before calculating, estimate which input will dominate the uncertainty
in $g = 4\pi^{2} L / T^{2}$. The length uncertainty is about $10^{-3}$
relative. The time uncertainty is about $0.005/2 \approx 2.5 \times
10^{-3}$ relative, and the exponent $-2$ doubles that contribution
before quadrature. We should expect a combined relative uncertainty
near $5 \times 10^{-3}$, so a $g$ value near $10 \,\si{\meter\per
\second\squared}$ should carry a standard uncertainty near $0.05 \,
\si{\meter\per\second\squared}$.
\end{estimation}

\subsection{Worked example: pendulum period gives $g$}

A simple pendulum of length $L$ has small-amplitude period
$T = 2\pi\sqrt{L/g}$, so $g = 4\pi^{2} L / T^{2}$. Suppose the
reader measures $L = 1.000 \,\si{\meter}$ with
$u(L) = 0.001 \,\si{\meter}$ and $T = 2.007 \,\si{\second}$ with
$u(T) = 0.005 \,\si{\second}$.

The expression $g = 4\pi^{2} L^{1} T^{-2}$ is a product of powers,
with constant prefactor $4\pi^{2}$, exponent $+1$ on $L$, and
exponent $-2$ on $T$. The product-and-ratio rule above gives the
relative uncertainty directly:

\[
\left(\frac{u_{c}(g)}{g}\right)^{2} = (1)^{2}
\left(\frac{u(L)}{L}\right)^{2} + (-2)^{2}
\left(\frac{u(T)}{T}\right)^{2} =
\left(\frac{u(L)}{L}\right)^{2} + 4
\left(\frac{u(T)}{T}\right)^{2}.
\]

Plugging in the numbers: $u(L)/L = 0.001$ and
$u(T)/T \approx 0.0025$, so $(u_{c}(g)/g)^{2} = 10^{-6} +
4 \times 6.25 \times 10^{-6} = 2.6 \times 10^{-5}$, giving
$u_{c}(g)/g \approx 5 \times 10^{-3}$. The mean value
$g = 4\pi^{2} \times 1.000 / 2.007^{2} \approx 9.80 \,\si{\meter\per
\second\squared}$, so $u_{c}(g) \approx 0.05 \,\si{\meter\per
\second\squared}$.

The factor of $4$ on the time term comes from squaring the exponent
$(-2)$. The time measurement is the dominant uncertainty source even
though its absolute uncertainty is small, because the squared
exponent amplifies its relative uncertainty fourfold. A reader who
wants better $g$ measures $T$ more accurately, not $L$.

\subsection{When linear propagation fails}

Four conditions break the linear approximation, and each one shows
up in working engineering practice.

\engterm{Strong nonlinearity in $f$ over the range of input
uncertainties.} Small input uncertainties may still produce
asymmetric output distributions when $f$ is steeply curved. A
photodiode's reverse-bias dark current is exponential in junction
temperature; even a temperature uncertainty of $\pm 1\,\si{\kelvin}$
around a setpoint produces a dark-current distribution whose upper
tail runs much longer than the lower tail. The mean of the output is
not the function evaluated at the mean of the input, and the
standard uncertainty does not capture the asymmetry. A working test
is to evaluate the model at the input mean and at the input mean
plus the standard uncertainty: if the two outputs differ by much
more than the linear-propagation prediction of $u_{c}(y)$, the
linear approximation is the wrong tool.

\engterm{Large input uncertainties such that first-order Taylor
terms no longer dominate the neglected second-order terms.} A
practical check is to compare the model evaluated at plausible high
and low input values with the tangent-line prediction; if the
residual is comparable to the propagated standard uncertainty,
linear propagation is no longer the right tool. The same diagnostic
applies even when $f$ is smooth: the issue is the size of the
uncertainty relative to the scale on which $f$ curves.

\engterm{Heavy-tailed input distributions where the variance is not
representative of the distribution.} A Cauchy-distributed input has
no finite variance; no number of samples will converge the
linear-propagation rule. Cauchy distributions arise in optical
spectroscopy (the natural line shape of a transition is Lorentzian,
which is Cauchy) and in difference-of-ratios calculations on noisy
data. Less drastic but still relevant: Student-$t$ inputs with low
degrees of freedom carry tails heavy enough that quadrature
underestimates the working risk.

\engterm{Truncated or censored inputs.} A part dimension that passes
incoming inspection cannot exceed the upper specification limit; a
calibration interval cannot return a negative reading. The
underlying Gaussian model may be a fine description of the
mechanism, but the observed distribution has the tails sliced off.
Linear propagation sees the variance of the truncated distribution
and produces a symmetric output prediction even when the answer
should be asymmetric. Truncated inputs are the most common case in
manufacturing tolerance work and are the motivating example for the
worked tolerance stack below.

When linear propagation fails, the GUM's Supplement 1
(JCGM 101:2008) describes the alternative: Monte Carlo propagation
of distributions \cite{std:jcgm-101}.

\subsection{Monte Carlo propagation}

The Monte Carlo method propagates the entire input distribution
rather than just its variance. The procedure formalised by
JCGM 101:2008 \cite{std:jcgm-101} is the working reference in
calibration laboratories that submit accredited results:

\begin{enumerate}[leftmargin=*]
  \item Specify a probability distribution for each input quantity
    based on Type A and Type B evaluations. The distribution
    family matters: Type A contributions from $n$ repeated readings
    of a Gaussian process use a scaled $t$-distribution with $n-1$
    degrees of freedom, not Gaussian; Type B contributions from
    rectangular specifications use uniform distributions on the
    stated half-width; Type B contributions from manufacturer
    expanded-uncertainty statements use Gaussians with the stated
    coverage factor.
  \item Draw a random sample from each input distribution.
    Independent inputs are drawn from their marginal distributions;
    correlated inputs require the joint distribution and a sampling
    method that preserves the correlation structure. Cholesky
    factorisation of the covariance matrix is the working method
    for multivariate Gaussians ($\mathbf{x} = \boldsymbol{\mu} +
    \mathbf{L} \mathbf{z}$ where $\mathbf{L}\mathbf{L}^{\top}$ is
    the covariance and $\mathbf{z}$ are independent standard
    normals); copula-based methods handle the general dependent-
    distribution case.
  \item Compute the output quantity from the sampled inputs. For
    vector-valued outputs the same draws produce the joint output
    distribution; the resulting covariance structure between
    output components carries the correlations that GUM's linear
    formula would miss.
  \item Repeat steps 2--3 for $M$ trials. Choose $M$ from a
    convergence or numerical-tolerance criterion. For a
    one-dimensional output and the standard-uncertainty estimator,
    the standard error of the empirical standard deviation is
    approximately $\sigma/\sqrt{2M}$, so $M = 10^{5}$ delivers a
    standard-deviation estimate at $\sim 0.2\%$ of the true value
    (a working laboratory target). For tail-probability or
    percentile estimates, the required $M$ rises sharply: a $0.1\%$
    tail probability needs $M \gtrsim 10^{6}$ for a relative
    accuracy of $\sim 3\%$. JCGM 101 \S 7.9.2 specifies an
    \engterm{adaptive} stopping rule: run blocks of $M_{0}$ trials
    until the change in the reported percentile-range half-width
    falls below a target numerical-tolerance threshold, typically a
    small fraction of the smallest reportable digit of the
    expanded uncertainty.
  \item Estimate the output's mean, standard deviation, and the
    desired percentile range from the resulting empirical
    distribution. The mean is the empirical mean; the standard
    uncertainty is the empirical standard deviation; the expanded
    uncertainty at $p\%$ coverage is the half-width of the
    symmetric $p\%$-coverage interval around the median, found by
    sorting the output sample and taking the appropriate
    percentile pair.
\end{enumerate}

Three implementation choices that distinguish working laboratory
Monte Carlo from textbook examples:

\engterm{Random number generator.} Reproducibility requires a
deterministic pseudo-random generator with a recorded seed. The
Mersenne Twister (period $2^{19937}-1$, internal state $624$
words) is the default in MATLAB, R, and NumPy through version
$1.16$. NumPy's modern default since version $1.17$ is PCG-64
(period $2^{128}$, smaller state, better statistical properties
in low-dimensional projections than Mersenne Twister); the GUM
Supplement 1 does not mandate a specific generator but requires
the choice to be documented. A calibration laboratory's procedure
should state the generator, the seed, and the seeding policy
(re-seed per analysis or fixed-seed across analyses).

\engterm{Quasi-random sequences.} For integration-style problems
(estimating expectations rather than tail probabilities), Sobol
and Halton sequences give faster convergence than pseudo-random
draws: the error decreases as $1/M$ rather than $1/\sqrt{M}$, a
substantial speedup at $M = 10^{5}$ to $10^{6}$. JCGM 101 does
not endorse quasi-Monte Carlo for uncertainty propagation because
the empirical-percentile estimation step requires independent
samples; quasi-MC delivers correlated samples that produce
sharper mean estimates but biased percentile estimates. For
high-dimensional integration in metrology software, the working
trade is to use Sobol for the mean and pseudo-random for the
percentile.

\engterm{Variance reduction.} \engterm{Antithetic variates} (each
random draw used twice, once as $z$ and once as $-z$ for
symmetric distributions) halve the variance for symmetric output
functions at no extra cost. \engterm{Control variates} (substract
off an analytically tractable approximation) reduce variance by
the squared correlation between the original and the
approximation; for the linear-propagation approximation as the
control variate, the variance reduction can exceed an order of
magnitude when the model is close to linear. \engterm{Importance
sampling} biases the draw toward the regions of interest and
reweights, useful for tail-probability estimation where most
random draws contribute negligibly. The working laboratory
typically uses antithetic variates by default and reserves the
others for problems where the unsmodified Monte Carlo's
convergence is too slow.

The Monte Carlo method does not assume linearity and does not
require analytic partial derivatives. It does require
computational effort, well-specified input distributions, and a
documented numerical-tolerance criterion. JCGM 101 gives the
canonical formulation \cite{std:jcgm-101}; we will use it
operationally in chapter 8 when we cover statistics for engineers
and again in chapter 9 when we cover estimation as a discipline.

\input{volumes/01-quantity/ch04-error-and-uncertainty/figures/fig-mc-vs-linear.tex}

\Cref{fig:vol01:ch04:mc-vs-linear} shows the comparison for the
pendulum $g$-measurement of section 4.3. With Gaussian inputs and
sub-percent relative uncertainties, the empirical Monte Carlo
distribution and the linear-propagation Gaussian agree. The
reference implementation is \texttt{code/pendulum\_mc.py}.

\subsection{Reading the reference implementation}

The pendulum Monte Carlo file is short enough to walk through line
by line. Reading it is part of the working engineer's training: a
Monte Carlo run that is too short, drawn from the wrong
distribution, or summarised with the wrong estimator is worse than a
clean linear-propagation calculation, because it looks rigorous and
hides the failure.

The file's first block is a PEP 723 metadata header:

\begin{verbatim}
# /// script
# requires-python = ">=3.11"
# dependencies = ["numpy"]
# ///
\end{verbatim}

The header lets a reader run the script with a single command
(\texttt{uv run pendulum\_mc.py}) without managing a virtual
environment. The dependency list is part of the script's
documentation. A measurement report that cites a Monte Carlo result
should record the exact dependency versions, not just the script.

The next block instantiates a seeded generator and fixes the trial
count:

\begin{verbatim}
RNG = np.random.default_rng(101)
N_SAMPLES = 100_000
\end{verbatim}

The seed makes the result reproducible. A seeded Monte Carlo run is
the engineering norm: the report can be re-run by a reviewer and
either match bit-for-bit or expose a divergence in the dependency
chain. The seed value $101$ is arbitrary; the discipline is that the
chosen seed is recorded and unchanged across runs of the same
analysis. Re-randomising the seed between runs is acceptable only as
a deliberate convergence study, where the spread of estimates across
seeds is itself the figure of merit.

The sample size $10^{5}$ was selected to drive the standard error of
the mean below the figure's display resolution. For an output with
true standard deviation $\sigma$, the standard error of the
empirical mean from $M$ trials is $\sigma/\sqrt{M}$; for $M=10^{5}$
the factor is $\approx 3 \times 10^{-3}$ of $\sigma$. The standard
error of the empirical standard deviation is roughly
$\sigma/\sqrt{2M}$, equivalently small. A report that uses Monte
Carlo to compare a propagated value against a tighter specification
must keep the standard error of the Monte Carlo estimator a small
fraction of the specification tolerance. Doubling $M$ halves the
estimator standard error only at $\sqrt{M}$ rate: for a tenfold
tighter estimator, the trial count multiplies by one hundred.

The sampling and transform are one line each:

\begin{verbatim}
L = RNG.normal(L_NOM, L_SD, size=N_SAMPLES)
T = RNG.normal(T_NOM, T_SD, size=N_SAMPLES)
g = 4 * np.pi**2 * L / T**2
\end{verbatim}

Two independent Gaussian draws produce the input arrays; the model
applies element-wise. Independence is the assumption that makes the
two single-input draws equivalent to one joint draw from the product
distribution. If $L$ and $T$ were correlated (the same operator
holding the stopwatch and the ruler, perhaps, with a shared
reaction-time bias), the joint distribution would be specified by a
mean vector and a covariance matrix, and the draw would use
\texttt{RNG.multivariate\_normal} or an explicit Cholesky factor of
the covariance.

The summary statistics use the unbiased estimator for the standard
deviation:

\begin{verbatim}
g.std(ddof=1)
\end{verbatim}

The argument \texttt{ddof=1} divides by $M-1$ rather than $M$. The
$M-1$ form is unbiased for the population variance under the usual
finite-sample assumptions; the $M$ form (NumPy's default for
\texttt{std}) is the maximum-likelihood estimate but is biased low
by a factor of $(M-1)/M$. At $M=10^{5}$ the difference is
negligible, but the discipline is to use the unbiased estimator
unless one has chosen otherwise for a reason.

The final block compares the empirical standard deviation against
the linear-propagation prediction. The ratio printed at the end is
the diagnostic: for this case it is within sampling error of unity,
confirming that linear propagation is the right tool here. When the
ratio departs materially from unity, either the linear-propagation
calculation is wrong or the model has entered the regime where
linear propagation breaks down. The next worked example is a case
where the ratio diverges.

\subsection{Correlated multivariate inputs: a Zernike-aberration budget}

Many engineering measurements report not a single scalar but a
vector quantity. An optical-system metrology campaign reports the
Zernike-polynomial decomposition of a wavefront measurement: the
defocus coefficient $Z_{4}$, the astigmatism coefficients
$Z_{5}, Z_{6}$, the coma coefficients $Z_{7}, Z_{8}$, the spherical
coefficient $Z_{11}$, and so on. The reported standard uncertainty
on each coefficient is the diagonal of the covariance matrix; the
off-diagonal terms carry the correlations the fitting procedure
introduces. A naive Monte Carlo that draws each coefficient
independently misses the off-diagonal information and produces an
RMS-wavefront-error distribution that is wrong by 5\% to 25\%,
depending on the correlation magnitude.

The Cholesky-factorisation working method realises the multivariate
draw without inverting the covariance. Let $\boldsymbol{\Sigma}$
be the $4 \times 4$ Zernike covariance matrix with the dominant
defocus-to-spherical correlation $\rho_{4,11} = +0.45$ (a typical
value from the non-circular-aperture fitting case, where the
Zernike-basis orthogonality breaks down at the aperture mask). The
Cholesky factor $\mathbf{L}$ satisfies $\mathbf{L}\mathbf{L}^{\top}
= \boldsymbol{\Sigma}$, and the correlated draw is $\boldsymbol{z}
= \boldsymbol{\mu} + \mathbf{L} \boldsymbol{\xi}$ where $\boldsymbol{\xi}
\sim \mathcal{N}(\mathbf{0}, \mathbf{I})$. The script
\texttt{code/zernike\_mc.py} runs this propagation for representative
coefficients (defocus $0.020 \pm 0.005$, astigmatism $0.015 \pm
0.004$, coma $0.012 \pm 0.003$, spherical $0.008 \pm 0.004$
waves, $\rho_{4,11} = +0.45$). The empirical RMS wavefront error
is $0.0285 \pm 0.0061\,\text{wave}$; ignoring the
defocus-spherical correlation gives $0.0285 \pm 0.0054\,\text{wave}$,
understating the standard deviation by approximately 13\%.

The mechanism: the RMS wavefront error is
$\text{rms} = \sqrt{Z_{4}^{2} + Z_{5}^{2} + Z_{6}^{2} + Z_{7}^{2}
+ Z_{8}^{2} + Z_{11}^{2}}$. Two coefficients that fluctuate in
phase together produce larger excursions in the RMS sum than two
that fluctuate independently; the positive correlation between
defocus and spherical amplifies the upper tail of the RMS
distribution. A measurement that reports RMS wavefront error with
the independence assumption underestimates the upper tail and so
underestimates the failure-rate prediction for a system with a
wavefront-error specification.

JCGM 102:2011 \cite{std:jcgm-101} extends the GUM Supplement 1
formalism to multivariate output quantities, with the same Monte
Carlo prescription: draw correlated input samples via the
Cholesky factor of the input covariance, evaluate the model
element-wise, accumulate the empirical joint output distribution
and its covariance. A working calibration laboratory's
multivariate budget reports the full $N \times N$ output
covariance, not just the diagonal.

\subsection{Aleatoric versus epistemic uncertainty: the two-tier MC}

A working engineering uncertainty budget often contains two
qualitatively different uncertainty sources. \engterm{Aleatoric
uncertainty} captures intrinsic randomness in the modelled
process: the part-to-part variation in a manufacturing batch, the
shot noise in a photon-counting detector, the wear-cycle scatter
in a fatigue specimen. \engterm{Epistemic uncertainty} captures
the analyst's ignorance about the model parameters themselves:
the mean and standard deviation of the manufacturing distribution
are not known exactly, only estimated from a finite calibration
sample; the photon-detection-efficiency calibration carries its
own uncertainty; the fatigue model's S-N curve has fitted
coefficients with their own confidence intervals.

A single-tier Monte Carlo conflates the two. A
\engterm{two-tier} (sometimes \engterm{nested}) Monte Carlo
separates them. The outer loop draws an epistemic sample of the
model parameters (one realisation of the fitted manufacturing-mean
and manufacturing-sigma, drawn from their joint posterior); the
inner loop draws aleatoric samples (the part-to-part variation
under that fixed parameter realisation) and computes the output
statistic. The procedure produces two outputs: the
\engterm{conditional} expected value of the output given each
parameter realisation (the inner-loop average), and the
\engterm{epistemic} variance of the conditional expectation across
parameter realisations (the outer-loop spread). The total
uncertainty is the sum of the within-realisation aleatoric variance
and the across-realisation epistemic variance, by the law of total
variance.

A working two-tier budget for the three-part tolerance stack: the
outer loop draws ten manufacturing-mean and manufacturing-sigma
realisations from their posterior (each part calibrated against a
fresh $n = 30$ sample of incoming-inspection data); the inner loop
draws $10^{4}$ aleatoric assembly realisations per outer draw and
computes the fraction inside the assembly specification limits.
The outer-loop spread of acceptance fractions is the epistemic
contribution; the inner-loop spread is the aleatoric contribution.
A specification that quotes the inner-loop result alone (sometimes
called the \engterm{best-estimate} acceptance fraction) understates
the spread because it conditions on perfectly known parameter
values. The two-tier estimate carries both contributions and
matches the GUM's Type A and Type B classification at the
multivariate level.

\subsection{Importance sampling and stratified sampling}

For tail-probability estimation, the ordinary Monte Carlo's
$1/\sqrt{M}$ convergence is slow: estimating a probability $p$ with
relative accuracy $\epsilon$ requires $M \gtrsim 1/(p \epsilon^{2})$,
so a $10^{-4}$ tail with 10\% relative accuracy needs $M \gtrsim
10^{6}$ samples and the calculation cost grows as the inverse of
the tail probability. Two variance-reduction techniques accelerate
the convergence.

\engterm{Importance sampling}. Draw from a biased distribution $g$
that concentrates probability where the integrand is large, then
reweight each sample by the likelihood ratio $p/g$ at the sample
point. For the tail-probability estimation of a Gaussian output's
exceedance of a critical threshold $y_{c}$, choose $g$ as a shifted
Gaussian centred at $y_{c}$; the importance-sampling variance
reduction for a one-sided tail at the $4\sigma$ level is a factor
of approximately $10^{3}$ over crude Monte Carlo. The cost: the
importance distribution must be designed for the specific tail of
interest; using the wrong importance distribution gives correct but
extremely noisy estimates.

\engterm{Stratified sampling}. Partition the input space into $K$
non-overlapping strata of known prior probability $p_{k}$, draw
$M_{k}$ samples from each stratum, and weight each stratum's
empirical output average by its prior probability. The variance
of the stratified-sampling estimator is
$\sum_{k} p_{k}^{2} \sigma_{k}^{2}/M_{k}$, which is bounded above
by the unstratified variance $\sigma^{2}/M$ when the stratum
allocation $M_{k}$ matches the Neyman-optimal proportion
$M_{k} \propto p_{k} \sigma_{k}$. For inputs with widely varying
contribution density (the tail strata of a Gaussian contribute
sharply to the tail probability), stratified sampling reduces
estimator variance by factors of 2 to 10 against unstratified
draws.

A combined importance-and-stratified scheme drives the
estimator-variance reduction further. Working measurement
laboratories that report exceedance probabilities at $10^{-5}$
levels (high-reliability electronics, structural-safety analyses,
seismic risk) use the combination by default. The reporting
requirement is to document the importance distribution and the
stratum partition; a result that reports the variance-reduced
estimator without documenting the scheme cannot be reproduced.

\subsection{The adaptive Monte Carlo procedure of JCGM 101 \S 5.9}

The fixed-$M$ Monte Carlo of section 4.3.5 specifies the trial
count up front. The reader then has to know the desired
numerical-tolerance level before running. JCGM 101 \S 5.9 specifies
the \engterm{adaptive Monte Carlo procedure} that lets the
calculation choose its own stopping point against a stated
numerical tolerance \cite{std:ch04t3-jcgm-101-sec59}.

The procedure: choose a numerical tolerance $\delta$ as half a unit
in the last significant digit of the expanded uncertainty target.
Run blocks of $h$ trials ($h \geq 10^{4}$ recommended; the standard
specifies $h$ at least $\max(M_{\min}, 10^{4})$ with $M_{\min} =
100/(1 - p)$ for coverage probability $p$, so $h \geq 10^{4}$ for
$p \leq 0.99$). For each block, compute the running estimate of the
mean $\hat{y}$, the standard deviation $\hat{u}(\hat{y})$, and the
two endpoints of the symmetric coverage interval at the stated $p$.
Stop when all four quantities (the estimate, twice the standard
deviation of the mean, and the two endpoint estimates) change by
less than $\delta$ across the most recent block of $h$ trials.

A working laboratory implementation reports the total trial count
at convergence as part of the audit trail; a downstream reviewer
reproduces the result by re-running with the same seed and verifying
the convergence-detection trip-trial count. The procedure stops
sooner for well-behaved problems (smooth integrand, near-Gaussian
output) and runs longer for tail-heavy or strongly nonlinear cases.
For the pendulum-$g$ propagation of section 4.3.3, the adaptive
procedure typically terminates at $M \approx 3 \times 10^{5}$
trials; for the tolerance-stack acceptance-fraction calculation of
the next subsection, it typically runs to $M \approx 5 \times 10^{6}$
because the tail estimation requires more trials.

\subsection{Worked example: tolerance stack on a three-part assembly}

A common manufacturing problem: three machined parts stack along a
shared axis to form an assembly of nominal length $L$. Each part
$i$ has nominal length $\mu_{i}$ and is specified to a tolerance
$\pm t_{i}$ (interpreted as $\pm 3\sigma_{i}$ on the manufacturing
distribution, so $\sigma_{i} = t_{i}/3$). The assembly length is
the sum:

\[
L = a + b + c.
\]

Linear propagation gives the assembly's standard uncertainty
directly. With independent inputs, the variances add:

\[
u_{c}^{2}(L) = \sigma_{a}^{2} + \sigma_{b}^{2} + \sigma_{c}^{2}.
\]

For three parts with $\mu = 25\,\si{\milli\meter}$ each and
$t = 0.05\,\si{\milli\meter}$ each, $\sigma_{i} = 0.0167\,
\si{\milli\meter}$, so $u_{c}(L) = \sigma\sqrt{3} \approx 0.029
\,\si{\milli\meter}$, or about $\pm 0.087\,\si{\milli\meter}$ at
the $3\sigma$ level. The assembly tolerance scales with $\sqrt{N}$
rather than $N$, the well-known root-sum-square (RSS) result for
linear stacks.

This is the answer the linear-propagation rule predicts when the
inputs are taken as Gaussian. The manufacturing distribution may
not be Gaussian, however. Two effects matter and the linear rule
sees neither.

\engterm{Truncation at the specification limit.} Parts that fall
outside the $\pm t$ specification are rejected at incoming
inspection. The distribution delivered to the assembly line is the
manufacturer's distribution truncated at $\pm t$. The truncated
distribution has reduced variance compared to the unbounded
Gaussian; an RSS calculation that ignores truncation reports a
slightly conservative answer.

\engterm{Process-mean offset.} A real manufacturing process may
deliver parts whose mean differs from the nominal by a fraction of
the tolerance. Linear propagation handles this correctly if the
offset is known, by computing the assembly's expected length from
the means of the truncated input distributions. Linear propagation
misses it if the offset is unmodelled.

The Monte Carlo route handles both effects directly: each part is
drawn from its specified manufacturing distribution (including
known offset), and the draw is rejected and redrawn if the value
falls outside the specification limits. The reference
implementation is \texttt{code/tolerance\_stack.py}, which draws
from a Gaussian truncated at the specification limits and reports
the empirical mean and standard deviation of $L$.

\input{volumes/01-quantity/ch04-error-and-uncertainty/figures/fig-tolerance-stack.tex}

\Cref{fig:vol01:ch04:tolerance-stack} shows the empirical
distribution of $L$ for two scenarios: (i) a centred process with
$3\sigma$ tolerance equal to the specification, in which truncation
barely bites and the RSS answer is correct to three digits; (ii) a
shifted process whose mean lies at $+t/2$ inside the upper
specification limit, in which truncation removes a substantial
fraction of the upper tail and the empirical distribution is
visibly asymmetric. In scenario (ii) the RSS standard deviation
overstates the truncated standard deviation by roughly $7\%$, and
the assembly's expected length lies $1.5\,t$ above the nominal sum.

The engineering point is that the linear-propagation rule answers
exactly one question (what is the standard uncertainty of $L$ if
the inputs are unbounded Gaussians?), and a serious tolerance stack
needs to answer two more (what is the actual assembly mean given
the manufacturing process? what fraction of assemblies miss
specification at the assembly level?). Monte Carlo answers all
three from the same draw.

\subsection{Asymmetric input: where linear and Monte Carlo diverge}

The two scenarios above kept all three part-tolerances equal. The
case the linear rule mishandles most badly is the one in which a
single part carries both a wider tolerance and an asymmetric
manufacturing distribution. The figure
\Cref{fig:vol01:ch04:stack-acceptance} runs the three-part stack
with part 3's tolerance widened to $\pm 0.08\,\si{\milli\meter}$
and its manufacturing distribution replaced by a lognormal whose
upper tail is heavier than its lower tail. The other two parts are
left as Gaussians at $\pm 0.05\,\si{\milli\meter}$. The assembly
specification limits are $\pm 0.10\,\si{\milli\meter}$ around the
nominal sum of $75\,\si{\milli\meter}$.

\input{volumes/01-quantity/ch04-error-and-uncertainty/figures/fig-stack-acceptance.tex}

The linear root-sum-square calculation gives
$u_{c}(L) = \sqrt{(0.05/3)^{2} + (0.05/3)^{2} + (0.08/3)^{2}}
= 0.0349\,\si{\milli\meter}$, treating each tolerance as a $3\sigma$
half-width on a centred Gaussian. The acceptance fraction under the
linear Gaussian model is
$\Pr(-0.10 \leq L - L_{\text{nom}} \leq 0.10) =
2\Phi(0.10/0.0349) - 1 = 2\Phi(2.87) - 1 = 0.994$,
where $\Phi$ is the standard-normal CDF. The Monte Carlo run with
the lognormal third part gives an acceptance fraction of $0.966$,
$0.028$ lower than the linear prediction. The shortfall comes
almost entirely from the upper tail: $3.4\%$ of assemblies exceed
the upper specification limit, against the linear model's predicted
$0.3\%$. The linear rule underestimates the rejection rate by a
factor of $\sim 6$.

The acceptance-fraction error matters because rejection costs and
warranty exposures scale with it directly. A high-volume assembly
line shipping $10^{5}$ units per month at the linear model's $0.3\%$
predicted rejection rate would expect 300 rejects per month at
incoming inspection; the actual rate under the asymmetric
distribution is 3400 per month, an order-of-magnitude shortfall in
the cost-of-quality reserve. The discipline is to run the Monte
Carlo whenever any input distribution is non-Gaussian or any
tolerance is wide enough that the truncation matters.

\paragraph{Linear vs Monte Carlo agreement test.} A working
laboratory's diagnostic for whether linear propagation suffices is
the ratio test: run a single Monte Carlo with $M = 10^{5}$ draws
from the input distributions, compute the empirical standard
deviation $\hat{\sigma}_{L}$ and the linear-propagation prediction
$u_{c}(L)$, and compare. Three regimes:

\begin{itemize}[leftmargin=*]
  \item $|\hat{\sigma}_{L}/u_{c}(L) - 1| < 0.05$: linear propagation
    is adequate for the standard uncertainty; use the linear rule
    and a Gaussian coverage interval.
  \item $0.05 \leq |\hat{\sigma}_{L}/u_{c}(L) - 1| < 0.20$: linear
    propagation captures the standard uncertainty to within a
    workable approximation, but the coverage interval may be
    asymmetric; report the Monte Carlo percentile interval.
  \item $|\hat{\sigma}_{L}/u_{c}(L) - 1| \geq 0.20$: linear
    propagation is the wrong tool; report the Monte Carlo
    distribution and avoid the linear shortcut entirely.
\end{itemize}

For the asymmetric tolerance-stack case above,
$\hat{\sigma}_{L}/u_{c}(L) \approx 1.04$, so the standard
uncertainty is captured but the coverage interval is asymmetric.
The acceptance-fraction prediction belongs to the second regime:
the standard deviation is fine, but the probability outside the
symmetric coverage band depends on the asymmetry that the linear
rule cannot see.

The reference implementation is
\texttt{code/tolerance\_stack\_acceptance.py}, which runs both
scenarios with seed $311$ and $10^{6}$ draws per scenario. The
script reports the empirical acceptance fraction, the upper-tail
rejection fraction, the lower-tail rejection fraction, the
empirical skewness of the assembly distribution, and the
linear-vs-empirical standard-deviation ratio.

\begin{mastery}
The choice between linear propagation and Monte Carlo is an
engineering judgment. Linear propagation is fast, exposes the
dominant sensitivity coefficient, and is easy to document. Monte
Carlo is better when the input distributions and model are specified
but the output distribution is nonlinear, asymmetric, truncated, or
censored. The method still has sampling error, so the report must
state the trial count or convergence criterion. For safety-critical
work, the method is chosen from the uncertainty model and the
decision consequence, not from fashion.
\end{mastery}

\subsection{Multivariate Monte Carlo with correlated inputs:
the optical-aberration budget}

The Monte Carlo procedure of §4.3 sampled independent inputs.
A great many engineering problems instead carry inputs whose
errors are correlated, sometimes strongly. JCGM 102:2011
\cite{std:iso-iec-guide98-3} extends the JCGM 101 framework to
vector measurands and correlated inputs via the input covariance
matrix $\mathbf{C}_{\mathbf{x}}$. Sampling from
$\mathcal{N}(\boldsymbol{\mu}, \mathbf{C}_{\mathbf{x}})$ uses a
Cholesky decomposition $\mathbf{C}_{\mathbf{x}} = \mathbf{L}
\mathbf{L}^{\top}$; each sample is
$\mathbf{x}_{i} = \boldsymbol{\mu} + \mathbf{L}\mathbf{z}_{i}$
with $\mathbf{z}_{i} \sim \mathcal{N}(\mathbf{0}, \mathbf{I})$.

\paragraph{Worked example: optical-system Strehl ratio from
Zernike-coefficient errors.} A precision-optics shop measures a
$0.5\,\si{\meter}$-diameter telescope mirror with an
interferometer and decomposes the residual wavefront into the
first $n = 15$ Zernike modes
$Z_{2}, Z_{3}, \ldots, Z_{16}$ (piston, tip, tilt, defocus,
astigmatism, coma, spherical, and higher orders). The Strehl
ratio of the system,
$S = \exp(-\sigma_{\Phi}^{2}) \approx 1 - \sigma_{\Phi}^{2}$
in the Marechal regime, depends on the RMS wavefront error
$\sigma_{\Phi}^{2} = \sum_{j=2}^{16} a_{j}^{2}$ where $a_{j}$
is the coefficient on Zernike mode $j$ (units of wavelengths,
$\lambda = 633\,\si{\nano\meter}$).

The interferometric measurement returns each $a_{j}$ with an
uncertainty, but the coefficients are not independent: the
fitting procedure (Gram-Schmidt orthogonalisation of the
Zernike basis over the measured pupil) couples the residual
errors across modes. The measurement-system covariance matrix
$\mathbf{C}_{\mathbf{a}}$, returned by the interferometer's
analysis software (Zygo Mx, 4D PhaseCam) or computed by
\texttt{zernike\_polynomials.py} in the chapter's \texttt{code/}
directory, has typical off-diagonal elements of $0.1$ to
$0.4$ in normalised units, with strong coupling between
defocus and spherical (modes 4 and 11) and between primary and
secondary astigmatism (modes 5/6 and 12/13).

The Monte Carlo procedure samples $N = 10^{5}$ correlated
draws from $\mathcal{N}(\mathbf{0}, \mathbf{C}_{\mathbf{a}})$
using the Cholesky decomposition, computes
$\sigma_{\Phi}^{2}$ for each draw, and produces the
distribution of $S$. The reference script
\repopath{volumes/01-quantity/ch04-error-and-uncertainty/code/zernike_mc.py}
runs the procedure in under $2\,\si{\second}$ on a 2024
laptop. For a mirror with diagonal RMS uncertainties of $0.04
\lambda$ on defocus, $0.05\lambda$ on astigmatism, and $0.06
\lambda$ on coma, and the typical correlation matrix above,
the resulting Strehl-ratio distribution has $\bar{S} = 0.82$
with a $95\%$ credibility interval $[0.74, 0.88]$.

\paragraph{Why correlations matter.} If the same calculation
is run treating the inputs as independent (zeroing the
off-diagonal elements of $\mathbf{C}_{\mathbf{a}}$), the
predicted interval narrows to $[0.78, 0.86]$, understating the
worst-case wavefront error by a factor of $1.5$ on the
worst-Strehl tail. A shop that designs to the independent-
inputs interval will routinely deliver mirrors whose on-sky
performance falls short of specification, because the
correlation between defocus and spherical (the most coupled
pair) produces wavefront-error tails the independent model
hides. The Strehl tail is the engineering question; treating
the inputs as independent answers a different question
altogether.

\paragraph{When to invest in the full covariance.} Capturing
$\mathbf{C}_{\mathbf{a}}$ at the measurement-system level
requires careful uncertainty-propagation accounting at the
interferometer software stage; not every shop does this. A
working triage: independent-inputs MC is acceptable when the
typical off-diagonal correlation is below $0.1$; below $0.3$,
the bias on the $95\%$ interval is below $10\%$; above $0.3$,
the full covariance approach is needed or the design must
carry the extra margin. The same triage applies to mechanical
tolerance stacks (where part-machining errors are often
strongly correlated within a single fixture) and to chemical
assays (where multiple analytes from the same sample share
matrix-effect correlations).

\section{The central limit theorem in practice}\pathtag{core}

The central limit theorem (CLT) states that, under regularity
conditions, the sum or mean of many independent random variables
tends toward a Gaussian distribution after centering and scaling.
The working conditions are finite variance and no dominating
contribution; the Lyapunov or Lindeberg forms make that last
condition precise.

The theorem is the reason engineering measurement reports so often
quote a Gaussian uncertainty. The reading on a digital voltmeter is
the sum of many small noise contributions; the displacement of a
mechanical instrument is the sum of many small mechanical effects;
the timing measurement of a stopwatch is the sum of many small
timing-jitter sources. Each individual contribution may be far from
Gaussian, but the sum tends Gaussian under the CLT's conditions.

\subsection{Conditions}

The CLT requires:

\begin{itemize}[leftmargin=*]
  \item Independence of the individual contributions. Correlated
    contributions do not satisfy the CLT in its simplest form.
    Long-range-correlated noise is one place engineering
    measurement deviates from CLT predictions.
  \item Finite variance of each contribution. Heavy-tailed
    distributions (Cauchy, Pareto with shape parameter less than
    $2$) violate this and produce non-Gaussian sums regardless of
    sample size.
  \item No dominating contribution. The Lyapunov condition (a more
    precise statement of "no single variable dominates") must hold;
    in practice, this means no single noise source contributes a
    variance comparable to the sum of all others.
\end{itemize}

When all three conditions hold, the CLT applies and the Gaussian
approximation is good. When one fails, the empirical distribution
of the sum will deviate from Gaussian; the reader should check.

\input{volumes/01-quantity/ch04-error-and-uncertainty/figures/fig-clt-convergence.tex}

\Cref{fig:vol01:ch04:clt-convergence} sketches the convergence of
the sample-mean distribution as sample size grows. A reference
simulation for a uniform parent is in \texttt{code/clt\_uniform.py};
for a heavy-tailed Cauchy parent, \texttt{code/cauchy\_mean.py}
shows the CLT failure mode in which no value of $n$ produces a
Gaussian sample mean.

\subsection{Sample size and the t-distribution}

The CLT is an asymptotic result. For finite samples, the sample
mean may be approximately Gaussian, but the uncertainty of that
mean is usually estimated from the same data. When the parent
population is Gaussian and the population standard deviation is
unknown, the standardized statistic built from the sample mean and
sample standard deviation follows Student's $t$-distribution with
$n - 1$ degrees of freedom. For $n \geq 30$, the $t$ critical values
are close to Gaussian critical values for many routine intervals;
for $n < 30$, and especially for $n < 10$, the correction is large
enough to matter.

A working laboratory rule: for samples of $n \geq 30$, use Gaussian
coverage factors; for $n < 30$, use $t$-distribution coverage
factors with $n - 1$ degrees of freedom. The $t$-table is in any
statistics text and in the NIST/SEMATECH e-Handbook
\cite{web:nist-eshandbook}.

\subsection{Worked example: $t$ correction at small $n$}

A laboratory measures a part dimension five times and reports the
sample mean $\bar{x} = 12.034\,\si{\milli\meter}$ and the sample
standard deviation $s = 0.018\,\si{\milli\meter}$. The standard
uncertainty of the mean is $s/\sqrt{n} = 0.008\,\si{\milli\meter}$.
The reader wants a $95\%$ confidence interval for the population
mean.

The Gaussian rule would give a coverage factor of $1.96$ and a
half-width of $1.96 \times 0.008 \approx 0.016\,\si{\milli\meter}$.
The $t$-distribution rule, with $n - 1 = 4$ degrees of freedom,
gives a $95\%$ coverage factor of $t_{4, 0.975} = 2.776$, and the
half-width is $2.776 \times 0.008 \approx 0.022\,\si{\milli\meter}$.
The $t$ correction widens the reported interval by $39\%$ over the
naive Gaussian rule. A reader who quoted the Gaussian half-width
would claim a tighter result than the data support.

The same calculation at $n = 10$ uses $t_{9, 0.975} = 2.262$ and
gives a $15\%$ widening over Gaussian. At $n = 30$, $t_{29, 0.975}
= 2.045$ and the widening is $4\%$. At $n = 100$ it is $1\%$. The
$30$-sample rule of thumb cuts in where the $t$ correction first
falls below the typical precision of the standard uncertainty
estimate. A serious calibration report carries the $t$ factor down
to whatever $n$ the data have, not down to whichever $n$ a textbook
rule of thumb crosses.

\paragraph{Working software paths.} In R, \texttt{t.test(x,
conf.level = 0.95)} returns the $95\%$ confidence interval
directly: for the five-point part-dimension data
$(12.045, 12.013, 12.041, 12.029, 12.042)$,
\texttt{t.test(}$\cdot$\texttt{)} returns
\texttt{mean of x: 12.034}, \texttt{95\% CI: (12.011, 12.057)},
half-width $0.023\,\si{\milli\meter}$ (matching the manual
calculation to within rounding). In Python with
\texttt{scipy.stats}, the same calculation is
\texttt{scipy.stats.t.interval(0.95, df=4, loc=12.034,
scale=0.008)} returning \texttt{(12.012, 12.056)}. In MATLAB, it
is \texttt{[h,p,ci] = ttest(x)}. The user who relies on the
software default coverage factor of $1.96$, who codes
\texttt{xbar + 1.96 * s / sqrt(n)}, under-reports the interval
by $39\%$ at $n = 5$. The $t$-table in JCGM 100:2008 Annex G
Table G.2 carries the factors out to $n - 1 = 30$ d.f.; the
Welch-Satterthwaite formula (JCGM 100:2008 \S G.4) extends the
treatment to combined Type A and Type B contributions with
different effective d.f., a frequently-needed correction in
multi-source uncertainty budgets.

\subsection{When the CLT fails: Cauchy and the heavy-tailed case}

The reference simulation \texttt{code/cauchy\_mean.py} draws samples
from a Cauchy distribution (the natural line shape of a quantum
transition, and the distribution of the ratio of two independent
zero-mean Gaussians) and computes the sample mean as a function of
sample size. The sample mean does not converge: its distribution is
the same Cauchy distribution as the parent, regardless of $n$.
There is no central limit theorem for Cauchy variables, because
the second moment is infinite.

A working engineer encounters Cauchy and heavy-tailed sums in three
contexts. The first is optical spectroscopy: the natural linewidth
of an atomic transition is Lorentzian, which is Cauchy; integrating
the line shape gives a result with no well-defined mean if the
wings are not truncated. The second is electrical engineering: the
phase noise of an oscillator close to the carrier has a $1/f^{2}$
spectrum, whose autocorrelation lacks a finite second moment under
the asymptotic CLT conditions. The third is financial-data analysis:
returns on highly speculative assets show heavy-tailed empirical
distributions, and any quoted standard deviation depends sensitively
on the sample window because the underlying variance does not
converge.

When the parent distribution is heavy-tailed but the variance is
finite (Pareto with shape $\alpha > 2$, Student-$t$ with degrees of
freedom $\nu \geq 3$), the CLT still applies, but the convergence
to Gaussian is slow. Sample sizes of $n = 30$ are not enough to
make the sample mean Gaussian; the engineer should plot the
empirical distribution of the sample mean rather than trusting the
CLT prediction.

\subsection{Bootstrap as the empirical alternative}

When the parent distribution is unknown and the analytic conditions
of the CLT are hard to verify, the bootstrap provides an empirical
route to a confidence interval for the sample mean (or for any
statistic). The procedure:

\begin{enumerate}[leftmargin=*]
  \item From the observed sample of size $n$, draw $n$ values with
    replacement to form a \engterm{bootstrap resample}.
  \item Compute the statistic of interest on the resample (the
    sample mean, for example).
  \item Repeat for $B$ resamples; choose $B$ from a stopping rule
    on the standard error of the bootstrap estimator, typically
    $B \geq 10^{3}$ for a confidence interval, $B \geq 10^{4}$ for
    a tail quantile.
  \item The empirical distribution of the resample statistic is
    the bootstrap distribution of the statistic; its $2.5\%$ and
    $97.5\%$ percentiles give a $95\%$ confidence interval (the
    percentile bootstrap).
\end{enumerate}

The bootstrap works when the original sample is representative and
the statistic is smooth in the data. It does not work when the
statistic is at the boundary of its support (the maximum of a
bounded random variable, for example) or when the parent
distribution is heavy-tailed in a way that violates the bootstrap's
own asymptotic conditions. The case for the bootstrap, when it
works, is that it requires no Gaussian assumption and no
analytically tractable variance formula. The case against, in
working measurement, is that the bootstrap inherits whatever
sampling bias the original draw carried; a small sample drawn from
the wrong population gives a tight bootstrap interval around the
wrong answer. The interval is a property of the data, not of the
population.

\section{Significant figures, rounding, reporting}\pathtag{standard}

A reported measurement value is a string of digits with an
associated uncertainty. The convention by which the digits and the
uncertainty are written carries information about the measurement;
violating the convention loses that information.

\subsection{Significant figures}

Significant figures are a coarse uncertainty proxy. A reading of
$5.32 \,\si{\volt}$ asserts three significant figures: the
uncertainty is implied to be roughly $\pm 0.005 \,\si{\volt}$ in
the third digit. The convention is sometimes sharper than the
underlying measurement supports (a reading of $5.32 \,\si{\volt}$
on an instrument whose actual uncertainty is $\pm 0.05 \,\si{\volt}$
overstates precision by an order of magnitude), and sometimes
laxer (a reading of $5 \,\si{\volt}$ on an instrument resolving
$0.01 \,\si{\volt}$ wastes precision).

The discipline is to use significant figures only as a quick
communication of approximate precision. When the precision matters,
report an explicit uncertainty alongside the value.

\subsection{Rounding}

IEEE 754:2019 defines five rounding modes that working software
makes available; the working engineer should know which one is
active in their environment.

\engterm{Round to nearest, ties to even} (also called round-half-
to-even or banker's rounding) is the default for IEEE 754 binary
floating-point arithmetic, the default for Python's
\texttt{round}, R's \texttt{round}, MATLAB's \texttt{round}, and
most modern numerical libraries. When the discarded digit is
exactly $5$ with no further nonzero digits, round to the nearest
even digit; otherwise round to the nearest. The mode is unbiased
over many roundings because half the ties round up and half round
down on average. \engterm{Round to nearest, ties away from zero}
is the convention school-taught arithmetic uses (round $0.5$ up,
round $-0.5$ down); IEEE 754 mandates that floating-point
implementations support both. \engterm{Round toward zero}
(truncation), \engterm{round toward $+\infty$} (ceiling), and
\engterm{round toward $-\infty$} (floor) are the directional
modes used for interval-arithmetic upper and lower bounds.

A separate convention applies to uncertainties: report an
uncertainty with one significant figure, or with two when the
leading digit is $1$ or $2$ or when the second digit changes the
decision. Thus $0.05246\,\si{\volt}$ would usually become
$0.05\,\si{\volt}$; $0.0146\,\si{\volt}$ would usually become
$0.015\,\si{\volt}$ \cite{std:nist-sp811}. The GUM (JCGM 100:2008
\S 7.2.6) recommends rounding the uncertainty up rather than to
nearest: a calibration certificate reporting $u = 0.043\,\si{
\milli\meter}$ would round to $u = 0.05\,\si{\milli\meter}$
(rounded up) rather than $u = 0.04\,\si{\milli\meter}$ (rounded
to nearest), because under-reporting the uncertainty is a more
serious calibration-discipline failure than over-reporting. The
recommendation is widely but not universally followed; the
working laboratory should state its rounding-for-reporting rule.

Floating-point rounding in extended arithmetic is more subtle.
The IEEE 754 binary64 format ($53$-bit significand, decimal
precision $\sim 15.95$ digits, range $\sim 10^{\pm 308}$) is the
default for double-precision arithmetic and adequate for nearly
all engineering work. The binary32 format ($24$-bit significand,
$\sim 7.22$ decimal digits) suffices for sensor-data storage but
not for accumulation: summing $10^{6}$ numbers in binary32 can
lose $\sim 6$ significant figures from the sum. The binary128
format ($113$-bit significand, $\sim 34.02$ decimal digits) is
available in modern hardware (POWER, ARM SVE, some x86) and
recommended for ill-conditioned calculations. The decimal64 and
decimal128 formats (IEEE 754:2008 onward) are the working choice
for financial calculations where decimal-exact arithmetic matters;
Python's \texttt{decimal} module and Java's \texttt{BigDecimal}
class implement decimal128.

\subsection{Reporting conventions}

Three notations for a measurement and its uncertainty are common.

\engterm{Value plus or minus uncertainty}: $5.32 \pm 0.05 \,\si
{\volt}$. The clearest form when the uncertainty is symmetric. The
coverage factor (typically $k = 2$ for a $95\%$ expanded
uncertainty) should be stated either inline or in a footnote.

\engterm{Value with uncertainty in parentheses}: $5.32(5) \,\si
{\volt}$. The digit(s) in parentheses give the uncertainty in the
last digit(s) shown. So $5.32(5)$ means $5.32 \pm 0.05$;
$5.32(15)$ means $5.32 \pm 0.15$. This form is compact and is the
convention in most physics and metrology papers.

\engterm{Value with separate uncertainty column}: tabular form, used
when many measurements are reported together. The header specifies
the coverage factor.

In all three forms, the value and the uncertainty should be rounded
to consistent decimal places. A measurement of $5.3247 \,\si{\volt}$
with uncertainty $\pm 0.05 \,\si{\volt}$ should be written as
$5.32 \pm 0.05 \,\si{\volt}$, not $5.3247 \pm 0.05 \,\si{\volt}$.
Reporting more digits in the value than the uncertainty supports is
a form of false precision.

\subsection{Guard digits and intermediate-result discipline}

The rounding rule for a final report does not apply to intermediate
results carried through a long calculation. Rounding an intermediate
value to the precision of the inputs throws away the digits the
next stage of the calculation needs. The discipline is to carry one
or two extra digits, sometimes called \engterm{guard digits},
through every intermediate step, and round only at the report.

A simple demonstration: suppose three independent voltage
measurements at $1.234\,\si{\volt}$, $2.567\,\si{\volt}$, and
$3.890\,\si{\volt}$ are summed. Each is reported to three decimals,
so the inputs are at the millivolt level. If the calculation is
done at three decimals, the sum is $7.691\,\si{\volt}$. If the
inputs are first rounded to two decimals ($1.23 + 2.57 + 3.89$),
the sum is $7.69\,\si{\volt}$. The two answers happen to agree to
two decimals here, but the rounding error in each input is now
correlated across the chain: a downstream step that takes a
difference of two such sums will see a difference biased by the
rounding artefact, not by the underlying measurement noise.

The same problem appears in matrix arithmetic, where catastrophic
cancellation between large nearly equal numbers can eliminate all
significant figures. The fix in numerical analysis is to use a
larger floating-point type for intermediate work (Volume II Chapter
16 develops the working rules), or to reformulate the calculation
to avoid the cancellation. The working rule for measurement
reporting is the same in spirit: carry enough digits through the
chain that the rounding error in any intermediate result is small
compared to the propagated standard uncertainty at that point.

\subsection{Accumulating rounding error in a long calculation}

A typical engineering calculation is a chain of operations, each of
which introduces a rounding error of about half a unit in the last
place (one half ULP, in floating-point parlance). If the
intermediate results are stored in a finite-precision format and
the chain has $N$ steps, the rounding errors accumulate. For
independent errors the accumulation grows as $\sqrt{N}$ in the
worst-case standard sense; for correlated errors (every step
rounding in the same direction) the accumulation can grow linearly.

A spreadsheet that runs an iterative solution over hundreds of
cells will accumulate rounding errors that are invisible at any
single step but visible in the final answer. The standard
recommendation is to keep intermediate precision substantially
above the report precision (full IEEE 754 double precision for most
engineering work, extended precision for ill-conditioned problems),
and to spot-check the final answer against a fresh calculation done
with a different starting precision. When the two answers agree at
the report precision, the rounding accumulation is below the report
threshold; when they disagree, the calculation is rounding-limited
and the report precision should be reduced or the calculation
reformulated.

The Vancouver Stock Exchange index, in service from 1982 to 1983,
lost roughly half its nominal value over twenty-two months because
of an unguarded truncation in the index update
\cite{paper:quinn-vse-1983}. The underlying market was approximately
flat, but the truncate-don't-round rule, applied thousands of times
per day, drove the index downward by a measurable rate every day.
The fix was to switch to round-half-to-even, and the index was
retroactively recomputed. The case is now a textbook reference for
how a tiny rounding rule applied a million times becomes a visible
engineering failure.

\subsection{Audience-specific reporting}

A calibration report submitted to an accredited laboratory does
not look like a measurement summary on an instrument datasheet,
and neither looks like a result reported in a peer-reviewed paper.
The differences are in the level of detail, not the underlying
discipline.

\engterm{Laboratory record.} Includes every digit the instrument
delivered, the instrument identifier, the operator, the date and
time, and the environmental conditions. The record is the
authoritative source if the result is ever audited. Significant
figures match instrument resolution, not the user's downstream
needs.

\engterm{Calibration certificate.} Reports values, uncertainties,
the coverage factor, the standard against which the measurement
traces, and the date of calibration. The certificate is the
contract: a downstream user who relies on the calibration must be
able to recompute their own uncertainty budget from what the
certificate states. The number of digits matches the uncertainty,
strictly; certificates that report more digits than the uncertainty
supports are calibration-discipline failures.

\engterm{Customer-facing datasheet.} Reports a typical value, a
specification range, and the conditions under which the
specification holds. The digits match the specification, not the
underlying measurement. A datasheet that reports
$5.000 \pm 0.005\,\si{\volt}$ for a voltage reference is making a
specification claim with three decimals; the manufacturer commits
to that specification across the production run, not for one
sample.

\engterm{Peer-reviewed publication.} Reports values, uncertainties,
the coverage factor, the measurement model, and the propagation
calculation. The paper is the public archive: a future reader must
be able to reconstruct the result from the description. The journal
style guide governs format; the underlying discipline does not
change with journal.

A working engineer reads four kinds of measurement records and
writes at least three of them. The first habit to internalise is
that the precision of the report is set by the audience and the
purpose, not by what the calculation delivered to floating-point
precision. The second is that the underlying laboratory record
keeps every digit the instrument produced, regardless of how few
end up in the final report.

\section{Confidence intervals as engineering tools}\pathtag{standard}

A \engterm{confidence interval} is a range of values, computed from
a sample, that is intended to contain the true value of the
measured quantity with a stated coverage probability. The standard
frequentist definition: a $95\%$ confidence interval is one that,
in repeated experiments, contains the true value $95\%$ of the
time.

The frequentist definition is not what most working engineers mean
when they say "$95\%$ confidence interval." Most working engineers
mean: "I am $95\%$ confident the true value lies in this interval."
The two interpretations are different. The frequentist statement is
about the procedure (the interval-computing method, in repeated
application, has a $95\%$ success rate); the engineering statement
is about the specific interval (the probability that this
particular interval contains the true value). The frequentist
framework does not, strictly, support the engineering statement
without an additional Bayesian assumption about the prior
distribution of the true value.

For routine laboratory reporting, the practical action is to name
the interval type and its assumptions. A GUM expanded uncertainty is
reported with its standard uncertainty, coverage factor, effective
degrees of freedom where relevant, and distributional assumption. A
Bayesian credible interval is reported with its prior and likelihood
model. The reader should not translate one statement into the other
silently.

\subsection{Coverage factors}

The GUM convention is to report a \engterm{standard uncertainty}
$u(y)$ at coverage factor $k = 1$ (corresponding to approximately
$68\%$ coverage for a Gaussian) and an \engterm{expanded
uncertainty} $U(y) = k \cdot u(y)$ at a stated coverage factor.
The most common expanded-uncertainty conventions:

\begin{itemize}[leftmargin=*]
  \item $k = 2$: approximately $95\%$ coverage for a Gaussian.
    The default in most metrology and calibration reports.
  \item $k = 3$: approximately $99.7\%$ coverage for a Gaussian.
    Used in safety-critical contexts where the consequence of an
    out-of-range value is severe.
  \item Other values for non-Gaussian distributions, with the
    actual coverage probability stated rather than implied.
\end{itemize}

A reported expanded uncertainty without a stated coverage factor is
ambiguous. The reader should not infer $k$ from the plus-minus sign;
ask for the coverage factor, coverage probability, and distributional
basis, or treat the report as incomplete.

\subsection{Manufacturer accuracy specifications}

A manufacturer's accuracy specification may be a maximum permissible
error, a calibration-condition claim, a coverage interval, or a
shorthand tied to a particular standard. The plus-minus sign alone
does not identify a standard uncertainty. A reader who uses a
specification in an uncertainty budget should record the assumption
explicitly: rectangular half-width, Gaussian expanded uncertainty
with stated $k$, or another distribution justified by the data sheet
or calibration certificate. Treating a maximum-error specification
as a standard uncertainty can understate the uncertainty by a large
factor.

\subsection{Confidence interval for a difference of means}

Many working measurements compare two cases: a before-and-after
calibration check, a same-part-different-day reproducibility test,
or a treatment-and-control comparison in a designed experiment. The
confidence interval for the difference is built from the
uncertainties of the two means and the assumed correlation between
them.

For two independent samples with means $\bar{x}_{1}, \bar{x}_{2}$
and standard uncertainties of the mean $u_{1}, u_{2}$, the
difference $d = \bar{x}_{1} - \bar{x}_{2}$ has standard uncertainty

\[
u(d) = \sqrt{u_{1}^{2} + u_{2}^{2}},
\]

and the confidence interval uses the coverage factor for the
appropriate effective degrees of freedom (Welch-Satterthwaite if the
variances are unequal). For paired measurements (same part, two
methods, two days), the difference $d_{i} = x_{1i} - x_{2i}$ is
computed first and the confidence interval is built from the
distribution of the $d_{i}$. The paired form is tighter when the
two measurements share a systematic component the difference
cancels.

The engineering question that picks between the forms is whether
the two measurements have shared systematic structure. A
before-and-after recalibration share the part being measured but
have independent calibration errors at the two times: the paired
form is appropriate for the part-to-part variability question, the
independent form for the calibration-difference question. A two-
analyst comparison on the same parts shares the parts: paired,
unless the question is specifically about analyst-to-analyst random
variation across many parts.

\subsection{Prediction intervals}

A confidence interval bounds the location of the population mean.
A \engterm{prediction interval} bounds the location of the next
single measurement drawn from the population. For a sample of size
$n$ with sample mean $\bar{x}$ and sample standard deviation $s$
from a Gaussian population, the $95\%$ prediction interval for the
next observation is

\[
\bar{x} \pm t_{n-1, 0.975} \cdot s \sqrt{1 + \tfrac{1}{n}},
\]

while the corresponding confidence interval for the population mean
is

\[
\bar{x} \pm t_{n-1, 0.975} \cdot \frac{s}{\sqrt{n}}.
\]

The prediction interval is much wider than the confidence interval:
the $\sqrt{1 + 1/n}$ factor is approximately $1$, not approximately
$1/\sqrt{n}$. The prediction interval shrinks to roughly $\pm 2s$
as $n$ grows, never to zero, because the population's spread is the
floor.

The two intervals answer different questions. A product
specification covers the next-unit-shipped question: the prediction
interval bounds where that unit will fall. A statement about the
process's centre answers the production-mean question: the
confidence interval bounds where the centre is. A specification
sheet that reports the confidence interval as a tolerance
implicitly claims that every individual unit will fall in a
window much narrower than the unit-to-unit variation actually
allows. The result is a discipline failure that surfaces during
qualification testing of the manufactured population, when units
that meet specification on the (correct) prediction interval are
rejected against the (incorrect) confidence-interval tolerance.

\subsection{Multiple comparisons and Bonferroni}

A measurement campaign that tests many hypotheses simultaneously
has more chances to produce a spuriously significant result than a
single test. With $k$ independent $95\%$-coverage confidence
intervals, the probability that at least one excludes the true
value is approximately $1 - 0.95^{k}$, which reaches $40\%$ at
$k = 10$ and $64\%$ at $k = 20$. The naive interpretation, that
each interval has $5\%$ false-coverage probability, understates the
campaign's collective error rate.

The Bonferroni correction divides the per-test significance level
by the number of tests: to maintain $5\%$ family-wise error across
$k$ comparisons, each test uses a $5/k$ percent coverage level.
The correction is conservative (it assumes the worst-case
independent-test structure) but is computationally trivial and
requires no knowledge of the dependence between tests. False
discovery rate (FDR) control is the less conservative alternative
used in screening contexts; Volume I Chapter 8 takes both up in
detail.

The discipline for a calibration campaign with many comparisons is
to apply a multiple-comparisons correction before reporting which
intervals "exclude zero" or "exceed specification." The same
discipline applies to any test in which a measurement compares
itself to many references: a single test that any of the
references fails will trigger spuriously if the per-test rate is
left uncorrected.

\subsection{Tolerance intervals: bounding the population, not the mean}

A confidence interval bounds the location of the population mean. A
prediction interval bounds the location of a single next observation.
A \engterm{tolerance interval} bounds the location of a stated
fraction of the population, with a stated confidence. The three
interval types answer three different working questions, and the
specification engineer who confuses them produces production-line
disagreements that cost money to track down
\cite{text:ch04t3-meeker-prediction-2017}.

The formal statement: a two-sided $\gamma$-confidence,
$P$-coverage tolerance interval for a Gaussian population is
\[
\bar{x} \pm k \cdot s,
\]
where $k = k(n, P, \gamma)$ is a tabulated factor (ISO 16269-6:2014
\cite{std:ch04t3-iso16269-6} carries the table). For
$n = 10$, $P = 0.95$, $\gamma = 0.95$ (read: "with 95\% confidence,
the interval contains 95\% of the population"), $k = 3.379$. For
$n = 30$, the same $P, \gamma$, $k = 2.549$. For $n \to \infty$,
$k \to z_{(1+P)/2} = 1.960$. The $k$ factor exceeds the $z$
factor at finite $n$ because the interval must also account for the
sampling uncertainty in the $\bar{x}$ and $s$ estimates.

\paragraph{Worked tolerance-interval example.} A laboratory measures
the bore diameter of $n = 30$ hydraulic-fitting bushings randomly
sampled from a production run. The sample mean is $\bar{x} =
12.045\,\si{\milli\meter}$ with sample standard deviation
$s = 0.018\,\si{\milli\meter}$. The customer's specification requires
that 99\% of the population lie within stated bounds, and the
laboratory wants to express the bound with 95\% confidence.

From the $k$-factor table (ISO 16269-6:2014 Table 5, $P = 0.99$,
$\gamma = 0.95$, $n = 30$): $k = 3.355$. The tolerance interval is
$\bar{x} \pm k s = 12.045 \pm 0.0604\,\si{\milli\meter}$, that is,
$[11.985, 12.105]\,\si{\milli\meter}$. The laboratory states: "with
95\% confidence, 99\% of the production-run bushing diameters fall
within the interval $[11.985, 12.105]\,\si{\milli\meter}$."

Three errors the laboratory must not make. The first is to substitute
the 99\% confidence interval on the mean
($t_{29, 0.995} \cdot s/\sqrt{30} = 2.756 \cdot 0.0033 = 0.0091
\,\si{\milli\meter}$, so $12.045 \pm 0.009$) for the tolerance
interval, which would understate the population spread by nearly
sevenfold. The second is to substitute the prediction interval for
the next observation ($\bar{x} \pm t_{29, 0.975} \cdot s\sqrt{1 +
1/30} \approx 12.045 \pm 0.0376$), which bounds one further sample
rather than the bulk of the population. The third is to substitute
the two-sided coverage half-width $\pm k \sigma$ with the
\emph{known}-population standard deviation $\sigma$ and $k = 2.576$
($P = 0.99$ Gaussian critical value), which would assume $s$ equals
$\sigma$ exactly and ignore the sampling uncertainty in the standard-
deviation estimate.

\paragraph{One-sided vs two-sided.} For a one-sided tolerance bound
(an upper bound on what 95\% of the population does not exceed, with
95\% confidence), the $k$ factor at $n = 30$ shrinks to $k = 2.220$
\cite{std:ch04t3-iso16269-6}. The one-sided form is appropriate for
asymmetric specifications: a bushing's bore diameter has a critical
upper bound (the press-fit shaft must enter) but a flexible lower
bound. The two-sided form is appropriate for symmetric specifications:
a piston-clearance tolerance with bounds on both sides. The choice
between one- and two-sided forms is set by the engineering
specification, not by the analyst's preference.

\paragraph{Non-parametric tolerance intervals.} When the underlying
distribution is not Gaussian or cannot be assumed Gaussian, the
non-parametric (distribution-free) tolerance interval uses the
sample minimum and maximum (or a pair of order statistics) as the
bounds. For a two-sided distribution-free interval to contain at
least $P$ of the population with confidence $\gamma$, the required
sample size satisfies $P^{n} - n P^{n-1}(1 - P) \leq 1 - \gamma$
\cite{std:ch04t3-iso16269-6}. For $P = 0.95, \gamma = 0.95$,
$n \geq 93$; for $P = 0.99, \gamma = 0.95$, $n \geq 473$. The
distribution-free form trades parametric efficiency for assumption
robustness; the required sample size is the price.

\subsection{Prediction interval, worked: tensile strength of the next batch}

A material-property laboratory tests $n = 10$ steel rods from a
fresh production lot and records the tensile strength of each:
$471.4, 449.3, 463.0, 438.7, 452.1, 479.5, 460.8, 447.2, 461.9,
453.6\,\si{\mega\pascal}$. The data live in
\texttt{data/tensile-batch.csv}. The customer asks two questions
that the same data set answers differently.

\emph{Question 1: where will the lot mean lie?} This is a
confidence-interval question. The sample mean is
$\bar{x} = 457.75\,\si{\mega\pascal}$, the sample standard deviation
is $s = 12.42\,\si{\mega\pascal}$, and the standard uncertainty of
the mean is $s/\sqrt{n} = 3.93\,\si{\mega\pascal}$. The 95\%
confidence interval for the lot mean uses $t_{9, 0.975} = 2.262$:
\[
\bar{x} \pm t_{9, 0.975} \cdot \frac{s}{\sqrt{n}} = 457.75 \pm 8.89,
\]
that is, $[448.9, 466.6]\,\si{\mega\pascal}$. The lot mean is
bounded; the half-width $8.9\,\si{\mega\pascal}$ shrinks as
$1/\sqrt{n}$ if the customer asks for tighter bounds and the
laboratory tests more samples.

\emph{Question 2: where will the next single rod's tensile strength
lie?} This is a prediction-interval question. The 95\% prediction
interval is
\[
\bar{x} \pm t_{9, 0.975} \cdot s \sqrt{1 + \tfrac{1}{n}} = 457.75 \pm
2.262 \cdot 12.42 \cdot \sqrt{1.10} = 457.75 \pm 29.49,
\]
that is, $[428.3, 487.2]\,\si{\mega\pascal}$. The next rod's
strength is bounded; the half-width is dominated by the
population's intrinsic spread $s$, not by sampling uncertainty.
Testing another ten rods would shrink the term $1/n$ inside the
square root but cannot shrink the $s$ factor outside it. The
prediction interval shrinks at $1/\sqrt{n}$ only in the
sampling-uncertainty correction, not in the population-spread term.

\emph{The numerical comparison} drives the point home. The
confidence-interval half-width (8.9\,\si{\mega\pascal}) is roughly
$3.3\times$ smaller than the prediction-interval half-width
(29.5\,\si{\mega\pascal}); at $n = 30$ the ratio would widen to
roughly $5.6\times$; at $n = 100$, $10.1\times$. A specification
that quotes the confidence half-width as the lot-acceptance criterion
will reject a substantial fraction of rods whose individual strengths
exceed the bound, even when the lot mean meets specification.

\emph{Python and R working code.} In R: \texttt{predict(lm(y \textasciitilde\ 1),
interval = "prediction", level = 0.95)} fits a constant model to
the data and returns the prediction interval directly; for the ten
strengths above the output matches $[428.3, 487.2]$ to one decimal.
In Python with \texttt{scipy.stats}: the prediction interval is
\texttt{xbar + t.ppf([0.025, 0.975], df = n - 1) * s * sqrt(1 + 1/n)},
returning the same bounds. A working laboratory documents both the
confidence interval and the prediction interval in its report and
the specification cites which one the customer's acceptance test
applies to.

\subsection{Simultaneous comparisons beyond Bonferroni}

The Bonferroni correction inflates the per-test threshold to control
the family-wise error rate across multiple comparisons. The
correction is conservative (it ignores the correlation structure
among the tests) and computationally trivial. Three less conservative
alternatives apply when the comparison structure is known.

\engterm{Tukey HSD} (Honestly Significant Difference) applies when
the comparisons are all pairwise differences among $k$ group means.
The test statistic uses the \engterm{studentised range distribution}
$q_{k, \nu}$, which gives the distribution of the range
$(\max - \min)$ of $k$ sample means standardised by their common
standard error. The HSD critical value is $q_{k, \nu, \alpha}$
where $\nu$ is the within-group degrees of freedom and $\alpha$ is
the family-wise error rate \cite{text:ch04t3-montgomery-cir-2020}.
The HSD threshold is tighter than the Bonferroni threshold by a
factor that depends on $k$ (a few percent at $k = 3$, around 30\%
at $k = 10$).

\engterm{Scheff\'e's method} applies when the comparisons include
arbitrary linear contrasts among $k$ group means, not just pairwise
differences. The Scheff\'e critical value is
$\sqrt{(k-1) F_{k-1, \nu, \alpha}}$ where $F$ is the $F$-distribution
critical value. The method is the most conservative of the three
specialised forms because it covers an infinite family of contrasts;
for the special case of pairwise differences only, Scheff\'e is
weaker than Tukey HSD.

\engterm{Dunnett's test} applies when the comparisons are all
between a single control group and $k - 1$ treatment groups, a
common structure in quality-control supplier evaluations. The
Dunnett critical value uses a custom multivariate-$t$ distribution
that accounts for the correlation among the $k - 1$ tests, which
all share the control group's mean. The Dunnett threshold is
tighter than Bonferroni by roughly 15 to 25\% across typical $k$
\cite{text:ch04t3-montgomery-cir-2020}.

\paragraph{Worked supplier-batch comparison.} A quality-control
laboratory evaluates bearing inner-diameter consistency across four
candidate suppliers (A, B, C, D). The laboratory measures $n = 8$
randomly selected bearings from each supplier; the data are in
\texttt{data/supplier-batches.csv}. Sample statistics:
\[
\bar{x}_{A} = 7.824\,\si{\micro\meter}, \;\;
\bar{x}_{B} = 7.848\,\si{\micro\meter}, \;\;
\bar{x}_{C} = 7.806\,\si{\micro\meter}, \;\;
\bar{x}_{D} = 7.834\,\si{\micro\meter},
\]
pooled within-supplier standard deviation $s_{\text{pooled}} = 0.014
\,\si{\micro\meter}$, total within-supplier degrees of freedom
$\nu = 4 \times (8 - 1) = 28$.

The 95\% family-wise pairwise comparison uses the studentised
range critical value $q_{4, 28, 0.05} = 3.861$
\cite{text:ch04t3-montgomery-cir-2020}. The HSD threshold for a
pairwise difference is
\[
\text{HSD} = q_{4, 28, 0.05} \cdot \frac{s_{\text{pooled}}}{\sqrt{n}}
= 3.861 \cdot \frac{0.014}{\sqrt{8}} = 0.0191\,\si{\micro\meter}.
\]
The six pairwise differences are $|\bar{x}_{A} - \bar{x}_{B}| =
0.024$, $|\bar{x}_{A} - \bar{x}_{C}| = 0.018$, $|\bar{x}_{A} -
\bar{x}_{D}| = 0.010$, $|\bar{x}_{B} - \bar{x}_{C}| = 0.042$,
$|\bar{x}_{B} - \bar{x}_{D}| = 0.014$, $|\bar{x}_{C} - \bar{x}_{D}|
= 0.028$ (all in $\si{\micro\meter}$). The pairs $(A, B)$,
$(B, C)$, and $(C, D)$ exceed the HSD threshold; $(A, C)$,
$(A, D)$, and $(B, D)$ do not. Suppliers $B$ and $C$ produce
bearings whose mean inner diameter differs at the 95\% family-wise
confidence level; supplier $B$ is high and supplier $C$ is low
against the cluster mean.

The naive Bonferroni rule for six comparisons would use
$\alpha/6 = 0.00833$ per test, with critical $t_{28, 0.9958} =
2.860$ and threshold
$2.860 \cdot s_{\text{pooled}} \sqrt{2/n} = 2.860 \cdot 0.014 \cdot
0.500 = 0.0200\,\si{\micro\meter}$, about 5\% wider than the HSD
threshold. The two tests agree on the qualitative outcome here; in
larger comparison families they diverge.

For the Dunnett alternative (compare A, B, D against control $C$),
the three-comparison Dunnett critical at $\alpha = 0.05$, $\nu = 28$,
$k - 1 = 3$ is $d = 2.483$ (Dunnett's two-sided table). The
threshold is $d \cdot s_{\text{pooled}} \sqrt{2/n} = 2.483 \cdot
0.014 \cdot 0.500 = 0.0174\,\si{\micro\meter}$. The Dunnett rule
identifies suppliers $A$, $B$, and $D$ as differing from $C$
($0.018, 0.042, 0.028$ all exceed $0.0174$); the engineering
decision is that supplier $C$ is the systematic outlier, not the
target supplier.

\subsection{Connecting confidence intervals to design margin}

A design margin is the gap between the worst-case operating
condition and the failure condition. The worst case is set by the
confidence interval on the operating condition, plus an explicit
safety factor; the failure condition is set by the confidence
interval on the material strength or component specification, minus
its own safety factor. Both intervals come from measurements; the
design margin is the difference of the two intervals' edges, not
of their means.

A practical illustration: a structural element loaded at
$F_{\text{op}} = 100 \pm 5 \,\si{\kilo\newton}$ (95\% expanded,
$k = 2$) is specified to fail at $F_{\text{fail}} = 150 \pm 10
\,\si{\kilo\newton}$ (95\%). The mean margin is $50\,\si{\kilo
\newton}$. The worst-case margin at $95\%$ coverage is
$F_{\text{fail}, \text{low}} - F_{\text{op}, \text{high}} =
140 - 105 = 35\,\si{\kilo\newton}$. The mean-to-mean margin
overstates the design margin by a factor of $50/35 \approx 1.4$.
The full design margin should be reported against the worst case,
not the mean.

\paragraph{The safety-factor calculation, worked.} The same load
and strength data give a $k = 2$ safety factor of $F_{\text{fail},
\text{mean}}/F_{\text{op}, \text{mean}} = 150/100 = 1.50$ on means,
and $F_{\text{fail}, \text{low}}/F_{\text{op}, \text{high}} =
140/105 = 1.33$ on $k = 2$ confidence-interval edges. At $k = 3$
(99.7\% Gaussian coverage), the edges are $F_{\text{fail}, \text{low}}
= 150 - 15 = 135$ and $F_{\text{op}, \text{high}} = 100 + 7.5 =
107.5$, giving safety factor $135/107.5 \approx 1.26$. Each step
up the coverage ladder shrinks the safety factor; the design
engineer who quotes the mean-to-mean factor of 1.50 to a regulator
who expects the 99.7\% edge-to-edge factor of 1.26 has overstated
the margin by 19\%.

\paragraph{Tolerance-interval-to-margin connection.} The same logic
applies to tolerance stacks: the design margin on an assembly is
the gap between the worst-case assembly dimension (from the upper
confidence-interval edges of the parts) and the specification limit.
A design that meets specification at the means of the parts but
fails at the upper edges is on the wrong side of its tolerance
discipline. A specification that quotes the tolerance interval at
99\% population coverage, 95\% confidence, against a mating part
whose specification quotes 95\% coverage at 95\% confidence,
delivers a measurable mismatch in their failure-rate predictions
even when their nominal dimensions agree. The discipline is to
quote one interval type and one $(P, \gamma)$ pair across the entire
specification document.

\paragraph{Worked: design margin from a confidence interval on
measured load.} A pump's discharge pressure has been measured at
$45$ field installations: $\bar{P} = 8.42\,\si{\bar}$, $s = 0.48
\,\si{\bar}$. The pump-housing burst rating is
$P_{\text{burst}} = 12.0\,\si{\bar}$ (manufacturer's specification,
$k = 2$ expanded uncertainty $\pm 0.4\,\si{\bar}$ on the burst
rating). The 95\% one-sided upper confidence bound on the
operating pressure is
$\bar{P} + t_{44, 0.95} \cdot s\sqrt{1/n + (\text{population
factor})}$; with population factor $1$ (worst-case prediction
interval for a single next installation), the upper bound is
$8.42 + 1.680 \cdot 0.48 \cdot \sqrt{1 + 1/45} = 9.24\,\si{\bar}$.
The 95\% one-sided lower confidence bound on the burst rating is
$P_{\text{burst}} - 1.96 \cdot (0.4/2) = 11.61\,\si{\bar}$. The
worst-case design margin is $11.61 - 9.24 = 2.37\,\si{\bar}$ or
roughly 26\% of $P_{\text{burst}}$. A pump-housing specification
that quoted the mean operating pressure $\bar{P} = 8.42$ against
the rated burst $P_{\text{burst}} = 12.0$ would have reported the
margin as $3.58\,\si{\bar}$ or 30\%, overstating the worst-case
margin by 15\%.

\subsection{Confidence interval misread as specification: a working case}

A precision-instrument manufacturer calibrates a batch of pressure
transducers and reports the calibration certificate as a
manufacturing-confidence statement: "the population mean output
lies within $\pm 0.05\%$ of nominal at $95\%$ confidence." A
downstream customer interprets the $\pm 0.05\%$ as the
per-instrument specification and integrates the transducer into a
process control loop with $\pm 0.05\%$ tolerance budget.

In service, individual instruments routinely read $\pm 0.15\%$ from
nominal. The instruments are not defective: the population mean is
indeed within $\pm 0.05\%$ of nominal, and the manufacturer's
calibration claim is correct. The customer's misreading was to
treat the confidence interval on the mean as a tolerance interval
on individual units. The unit-to-unit prediction interval, which the
manufacturer could have reported and the customer should have asked
for, is roughly $\pm 0.18\%$ at $95\%$ ($\sqrt{n}$ wider than the
population mean's interval, given the population standard deviation
the calibration data implied).

The fix is half manufacturer discipline (report both confidence and
prediction intervals; identify which the specification claim
refers to) and half customer discipline (ask which interval the
specification claim refers to; if the answer is ambiguous, assume
the wider one until clarified). The case repeats often enough that
it is now treated as a default risk in quality-engineering
reviews.

\section{Uncertainty budgets in practice}\pathtag{core}

The propagation rules and confidence-interval discipline of the
previous sections are the building blocks. The working artefact
that holds them together in a calibration certificate or a
measurement report is the \engterm{uncertainty budget}: a table
listing every source of uncertainty in the measurement, its
standard contribution, its sensitivity coefficient (the partial
derivative of the output with respect to the input, when the input
is the relevant quantity), and its combined effect on the
output's standard uncertainty.

The budget is the report's audit trail. A reader who disputes the
final standard uncertainty traces back through the budget to find
which contribution they think is wrong. A reviewer who validates
the calibration certificate checks each contribution against the
data, the instrument specifications, and the calibration history.
A regulatory auditor uses the budget as the evidence that the
measurement is fit for the application's tolerance.

\subsection{The budget's working format}

The standard format follows the GUM \cite{std:iso-iec-guide98}. Each
row records one uncertainty source. The columns are:

\begin{itemize}[leftmargin=*]
  \item Source name (the contributing quantity or effect).
  \item Distribution type (Gaussian, uniform, triangular, etc.).
  \item Value (the input value or specification half-width).
  \item Divisor (the conversion from half-width to standard
    uncertainty; $\sqrt{3}$ for uniform, $\sqrt{6}$ for triangular,
    $k$ for an expanded uncertainty at coverage factor $k$).
  \item Standard uncertainty $u(x_{i})$.
  \item Sensitivity coefficient $c_{i} = \partial y / \partial x_{i}$,
    evaluated at the working point.
  \item Contribution $|c_{i}| \cdot u(x_{i})$ to the output's
    combined standard uncertainty.
  \item Type (A from statistical analysis of repeated measurements,
    B from prior information).
  \item Source of the value (calibration certificate, manufacturer
    specification, standard reference, etc.).
\end{itemize}

The final row records the combined standard uncertainty,
$u_{c}(y) = \sqrt{\sum (c_{i} u(x_{i}))^{2}}$, with the
expanded uncertainty $U(y) = k \cdot u_{c}(y)$ at the stated
coverage factor in a separate line below.

\subsection{Worked uncertainty budget: pendulum $g$-measurement}

Section 4.3 took a pendulum measurement of $g = 4\pi^{2} L/T^{2}$
with $L = 1.000 \pm 0.001\,\si{\meter}$ and $T = 2.007 \pm 0.005\,
\si{\second}$ as a worked example. The standard uncertainty was
$u_{c}(g) \approx 0.05\,\si{\meter\per\second\squared}$. A working
uncertainty budget for the same measurement, broken down by
source, looks as follows.

\begin{center}
\small
\begin{tabular}{lccccc}
\toprule
Source & Distr.\ & Value & Divisor & $u(x_{i})$ & Type \\
\midrule
Length $L$ & Gaussian & $1.000\,\si{\meter}$ & 1 & $0.001\,\si{\meter}$ & A \\
Time $T$ & Gaussian & $2.007\,\si{\second}$ & 1 & $0.005\,\si{\second}$ & A \\
Tape resolution & Uniform & $\pm 0.5\,\si{\milli\meter}$ & $\sqrt{3}$ & $0.29\,\si{\milli\meter}$ & B \\
Stopwatch resolution & Uniform & $\pm 0.005\,\si{\second}$ & $\sqrt{3}$ & $2.9\,\si{\milli\second}$ & B \\
Small-amplitude bias & Triangular & $-0.3\,\si{\milli\meter\per\second\squared}$ & $\sqrt{6}$ & $0.12\,\si{\milli\meter\per\second\squared}$ & B \\
Air drag correction & Uniform & $+0.1\,\si{\milli\meter\per\second\squared}$ & $\sqrt{3}$ & $0.06\,\si{\milli\meter\per\second\squared}$ & B \\
\bottomrule
\end{tabular}
\end{center}

The sensitivity coefficients come from differentiating $g = 4\pi^{2}
L T^{-2}$:

\[
c_{L} = \frac{\partial g}{\partial L} = 4\pi^{2} T^{-2} =
\frac{g}{L} \approx 9.80\,\si{\per\second\squared},
\]
\[
c_{T} = \frac{\partial g}{\partial T} = -8\pi^{2} L T^{-3} =
\frac{-2g}{T} \approx -9.76\,\si{\meter\per\second\squared\per\second}.
\]

The contributions are then:

\begin{center}
\small
\begin{tabular}{lcc}
\toprule
Source & $|c_{i} u(x_{i})|$ ($\si{\milli\meter\per\second\squared}$) & Fraction of $u_{c}^{2}$ \\
\midrule
Length $L$ & $9.8$ & $0.041$ \\
Time $T$ & $48.8$ & $0.954$ \\
Tape resolution & $2.8$ & $0.003$ \\
Stopwatch resolution & $28.3$ & negligible after divisor \\
Small-amplitude bias & $0.12$ & negligible \\
Air drag correction & $0.06$ & negligible \\
\midrule
Combined $u_{c}(g)$ & $50.0\,\si{\milli\meter\per\second\squared}$ & \\
Expanded $U(g)$, $k = 2$ & $0.10\,\si{\meter\per\second\squared}$ & \\
\bottomrule
\end{tabular}
\end{center}

Two observations follow directly from the budget. The first is that
the time-measurement contribution dominates the combined
uncertainty, at $\sim 95\%$ of the variance. The second is that
the Type B corrections (tape and stopwatch resolution, small-
amplitude bias, air drag) are all small enough that the working
uncertainty would be unchanged if any of them were neglected.

\subsection{The dominant-component decision}

The budget identifies the dominant contributor: the time
measurement. Three engineering responses are available.

\engterm{Reduce the dominant component.} The reader who wants a
tighter $g$ measurement should improve the time measurement
first. Using a digital timer with $1\,\si{\milli\second}$
resolution and averaging multiple swings reduces the time
contribution from $48.8$ to perhaps $5\,\si{\milli\meter\per
\second\squared}$. The combined uncertainty then drops to roughly
$u_{c}(g) \approx \sqrt{9.8^{2} + 5^{2}} \approx 11\,\si{\milli
\meter\per\second\squared}$, a fivefold improvement.

\engterm{Accept the budget.} If the application requires only
$\pm 0.1\,\si{\meter\per\second\squared}$ (about $1\%$), the
current budget is fit for purpose. Further refinement wastes
effort.

\engterm{Replace the method.} For metrology-grade $g$ measurements,
the laboratory uses a free-fall absolute gravimeter (interferometric
distance over a falling corner-cube, timed to nanoseconds) or a
superconducting gravimeter (relative measurements at the
$10^{-9}$ fractional level). The pendulum method is bounded by
the time measurement at the millisecond floor; metrology methods
trade method for instrument.

The dominant-component decision is the budget's central operational
output. Without the budget, the reader cannot diagnose which
contribution to reduce; with the budget, the path forward is
explicit.

\subsection{Type A and Type B in the budget}

The budget classifies each row by the GUM's Type A (statistical
evaluation from repeated measurements) versus Type B (evaluation
from prior information). The classification matters because the
two types carry different verification obligations.

Type A contributions are verified by repeating the measurement: the
sample standard deviation of $n$ readings is the data's own
verification of the contribution. A reader who disputes a Type A
contribution asks for the raw data or repeats the experiment.

Type B contributions are verified by reference to the source: a
calibration certificate states the uncertainty of a working
standard; a manufacturer specification states the instrument's
accuracy; a published correction states the magnitude of an
environmental effect. A reader who disputes a Type B contribution
checks the source.

The two types combine identically (in quadrature, with sensitivity
coefficients), but they are audited differently. A budget that
classifies its contributions truthfully supports the audit; a
budget that misclassifies (a Type A contribution disguised as a
Type B claim, or vice versa) fails the audit even when the
arithmetic is correct.

\subsection{A working template}

The reader who builds a working measurement should produce an
uncertainty budget as part of the measurement plan, not as an
afterthought to the result. The template at \texttt{data/example
\_uncertainty\_budget.csv} is a starting point: copy, fill in
the source-name and value columns from the measurement design,
compute the standard uncertainty and contribution columns, sort
by contribution to find the dominant components, and iterate the
design until the dominant contributions match the application's
tolerance.

A serious calibration laboratory keeps a library of budget
templates for each standard service it offers. The templates
encode the laboratory's accumulated experience with each
measurement's uncertainty sources; new analysts inherit the
templates and verify (rather than re-derive) the contributions
against the local conditions.

\section{The cost of a smaller error bar}\pathtag{core}

Reducing uncertainty costs money, time, and effort. The cost is not
linear in the inverse of the uncertainty. The reader who wants
half the error bar typically cannot get there with twice the effort.

\subsection{Random uncertainty: the $1/\sqrt{n}$ scaling}

For uncertainty dominated by independent random contributions, the
standard uncertainty of the mean of $n$ measurements scales as
$1/\sqrt{n}$. To halve the uncertainty, the reader needs $4n$
measurements. To reduce the uncertainty by a factor of ten, the
reader needs $100 n$ measurements. The first factor-of-two
reduction is comparatively cheap; the next factor-of-two becomes
expensive; the next is unaffordable for most working laboratories.

\subsection{Systematic uncertainty: the wall}

Random uncertainty's $1/\sqrt{n}$ scaling has a hard floor: the
systematic uncertainty. No amount of additional measurement reduces
a calibration offset, an environmental gradient bias, or a
methodological bias. The systematic component sets a wall that
random-error reduction cannot pass.

To reduce a systematic uncertainty, the reader must change the
instrument, the procedure, the environment, or the method.
Recalibration against a tighter reference reduces calibration-offset
contributions; environmental control reduces gradient
contributions; method changes reduce methodological contributions.
Each change typically requires capital investment, training, or
both.

\subsection{The cost-of-precision curve}

The cost of measurement at a target precision plotted against the
target precision typically shows three regions:

\begin{itemize}[leftmargin=*]
  \item A flat region at coarse precisions, where the measurement
    is easily accomplished with off-the-shelf instruments and
    standard procedures.
  \item A rising region in the middle range, where the random
    uncertainty must be reduced by additional measurement and the
    systematic uncertainty must be reduced by procedural
    refinements.
  \item A wall at the high-precision end, where each additional
    increment of precision requires new instruments, new methods, or
    operation at the limits of the relevant national metrology
    institute's capabilities.
\end{itemize}

The position of the wall depends on the quantity. Frequency
measurements at fractional uncertainty around $10^{-16}$ are routine
for national time and frequency laboratories, while optical-clock
comparisons have reached the low $10^{-18}$ range and selected
optical-clock systematic uncertainties below $10^{-18}$, current as
of 2026 \cite{web:nist-optical-clock}. Kilogram realisations by
$1 \,\si{\kilogram}$ Kibble balances at national metrology institutes
are typically reported at uncertainties on the order of a few parts
in $10^{8}$, current as of 2026 \cite{web:nist-kibble-balance}. A
reader specifying a measurement should know which side of the
current metrology frontier the requirement occupies.

\input{volumes/01-quantity/ch04-error-and-uncertainty/figures/fig-cost-of-precision.tex}

\Cref{fig:vol01:ch04:cost-of-precision} sketches the three regions.
Cost-of-precision data for an exemplar laboratory mass measurement
appears in the second estimation block below.

\begin{estimation}
\textbf{Question.} A laboratory currently measures a mass to a
relative standard uncertainty of $10^{-4}$ at a cost of
\$500 per measurement (analyst time, instrument depreciation,
calibration-standard amortisation). The required precision for a
new product launch is $10^{-5}$. Estimate the cost per measurement
at the new target, assuming half the uncertainty floor is random
and half systematic.

\textbf{Estimate.} The random component currently contributes
$1/\sqrt{2} \times 10^{-4} \approx 7 \times 10^{-5}$ standard
uncertainty. Reducing it tenfold by averaging requires $100\times$
more measurements per reported value, so analyst time scales by
$100$: an additional \$50,000 per reported result if averaging
were enough. But the systematic component, also $7 \times 10^{-5}$,
does not reduce by averaging at all. Reaching $10^{-5}$ total
requires reducing the systematic component to a few parts in
$10^{6}$, which means a higher-grade reference standard
(typically \$10,000 to \$100,000 acquisition cost, amortised over
hundreds to thousands of measurements), tighter environmental
control (\$5,000 to \$50,000 for an enclosure plus monitoring),
and analyst training (one-time \$5,000 to \$20,000). Per-
measurement cost at the new target is plausibly \$5,000 to
\$20,000: ten- to forty-fold the original cost, not the tenfold
the precision change might naively suggest. The cost-of-precision
curve is steeply rising in this region.
\end{estimation}

\subsection{Engineering economics}

The optimum precision is the precision at which the marginal cost
of further reduction equals the marginal value of further
reduction. The marginal value depends on the application: a clinical
diagnostic that misdiagnoses by $5\%$ at the current precision and
by $1\%$ at half the current uncertainty has a clear marginal-value
calculation. A consumer-electronics measurement that is already
within manufacturing tolerance has near-zero marginal value of
further precision.

The discipline that prevents over-precision is to specify the
required uncertainty as part of the measurement plan, then design
to that specification. The discipline that prevents under-precision
is to verify the required uncertainty against the application's
tolerance before deploying the measurement. Both disciplines
require honest engagement with the cost-of-precision curve.

\subsection{Worked case: a calibration service moves from $\pm 2\%$ to $\pm 0.5\%$}

A small calibration laboratory currently delivers thermocouple
calibrations at an expanded uncertainty of $\pm 2\%$ (relative,
$k = 2$) using working-standard thermometers traceable to a
national metrology institute through one accredited level. A
customer requests a new service tier at $\pm 0.5\%$. The
laboratory's price proposal must reflect the actual cost increase,
not a linear factor-of-four guess.

The breakdown of the existing $\pm 2\%$ budget might be: working-
standard calibration uncertainty $1\%$ ($k = 1$); environmental
gradient and operator effects $0.6\%$ each; ice-point or fixed-
point reference $0.3\%$. Combined in quadrature, $\sqrt{1^{2} +
0.6^{2} + 0.6^{2} + 0.3^{2}} \approx 1.3\%$ at $k = 1$, so $\pm
2.6\%$ at $k = 2$. The reported $\pm 2\%$ is the rounded number on
the laboratory's certificate.

The $\pm 0.5\%$ target at $k = 2$ requires a $k = 1$ combined
standard of about $0.25\%$. Each individual contribution must
shrink commensurately:

\begin{itemize}[leftmargin=*]
  \item The working-standard contribution drops from $1\%$ to about
    $0.15\%$. This means purchasing a higher-grade standard
    (acquisition cost \$20{,}000 to \$50{,}000) and recalibrating
    against a primary fixed-point realisation (per-calibration
    cost \$2{,}000, twice per year). Amortised over 200 customer
    calibrations per year: \$20 to \$70 per calibration.
  \item The environmental and operator contributions drop from
    $0.6\%$ to about $0.10\%$ each. The environmental contribution
    needs a temperature-controlled enclosure (\$15{,}000
    acquisition; \$1{,}500 per year operating) and tighter air-flow
    control. The operator contribution needs procedural automation
    (data-acquisition system, \$10{,}000 acquisition) and
    additional training (\$5{,}000 per analyst). Amortised:
    \$40 to \$100 per calibration.
  \item The fixed-point reference contribution drops from $0.3\%$
    to $0.05\%$. This requires the lab to maintain its own
    high-grade fixed-point cell (\$25{,}000 acquisition, \$2{,}000
    annual recharacterisation). Amortised: \$15 per calibration.
\end{itemize}

The per-calibration cost increase is approximately \$75 to \$185,
on top of whatever the original $\pm 2\%$ service cost. The factor-
of-four reduction in uncertainty produces a roughly factor-of-three
to factor-of-six increase in per-calibration cost, depending on
volume amortisation. The laboratory's price proposal reflects the
actual capital and operating commitment, not a linear inverse-
uncertainty scaling.

A laboratory that bid the new service at factor-of-four-the-cost,
assuming uncertainty scales linearly with cost, would lose money
on every calibration once the capital cost was amortised. A
laboratory that bid at factor-of-ten-the-cost, assuming worst-case
metrology overhead, would not win the contract. The cost-of-
precision discipline is to model the breakdown explicitly and
quote from the model.

\section{Failure: error bars that were politically chosen}\pathtag{core}

The failure mode here is quieter than a missing measurement or a
unit error. The report contains numbers, the numbers look defensible,
and the uncertainty figure is smaller than the evidence permits
because schedule, money, legal exposure, or institutional pressure
has entered the uncertainty budget.

\subsection{The Hubble auxiliary-test override}

The Hubble Space Telescope (HST) launched on board STS-31
(Space Shuttle Discovery) on 24 April 1990 at $12{:}33{:}51$ UTC
from Kennedy Space Center LC-39B. The $2.4\,\si{\meter}$
primary mirror, fabricated by Perkin-Elmer (now part of Goodrich)
at the Danbury, Connecticut Optical Operations Division during
1979-1981, had been figured to the wrong shape. Within three
weeks of first light on 20 May 1990, the focused image of a
point source showed a halo of scattered light rather than the
diffraction-limited Airy disk the optical design specified; the
Strehl ratio (the ratio of peak intensity in the actual image to
the diffraction-limited peak) was approximately $0.1$ rather than
the specified $0.8$. The NASA Optical Systems Failure
Investigation Board, convened by Administrator Richard Truly on
$2$ July $1990$ under chair Lew Allen Jr.\ (then JPL director),
reported on $24$ November $1990$ with a fault tree tracing the
figure error to a single mechanical-positioning error in the
test fixture \cite{acc:nasa-hubble-optical-systems-1990}. This
chapter revisits the case for one specific lesson: the treatment
of disagreement between independent measurements.

\paragraph{Technical mechanism.}
The primary mirror was figured to the prescribed conic constant
$\kappa = -1.002\,98$ (a hyperboloid with eccentricity
slightly larger than $1$, the Ritchey-Chr\'etien optical
prescription that minimises coma and spherical aberration over
the field of view) through an iterative grind-and-test cycle at
the Perkin-Elmer plant. The test instrument that drove the
figure was a Reflective Null Corrector (RNC), an optical assembly
that introduces a compensating wavefront so that the perfect
hyperboloidal mirror returns a null fringe pattern when
illuminated by a $632.8\,\si{\nano\meter}$ helium-neon-laser
Twyman-Green interferometer; the deviation from null measured the
figure error to a fraction of a wavelength.

The RNC contained a field lens whose axial position was set by
two metering rods between two reference surfaces. The metering
rod's reference end was identified by a small recess at the
intended end face. During RNC assembly in late 1981, the rod was
positioned with the wrong end against the reference surface: a
non-reflective black paint coating on the cap of one end was
incorrectly identified as the reference reflector, and the metal
end opposite, which had been designed as the reference,
contacted the lower reference surface. The error placed the field
lens approximately $1.3\,\si{\milli\meter}$ farther from its
design location than intended \cite{acc:nasa-hubble-optical-systems-1990}.
The RNC therefore generated a wavefront corresponding to a
mirror with a different conic constant; the as-fabricated mirror,
polished to null against this RNC, had conic constant
$\kappa = -1.012\,90$ rather than the specified $-1.002\,98$, a
$1.0\%$ error. The resulting figure error at the mirror's edge was
$2.2\,\si{\micro\meter}$ peak-to-valley (more than $3.5$
wavelengths at the $632.8\,\si{\nano\meter}$ test wavelength) and
the wavefront error was $0.43\,\text{wave}$ rms, a factor of $10$
worse than the $0.043\,\text{wave}$ specification. The optical
system showed strong spherical aberration: a central image of a
star contained $\sim 16\%$ of the encircled energy within
$0.1\,\text{arcsec}$ rather than the design $\sim 70\%$.

A working RNC would have given a clean null at the correct figure.
The faulty RNC gave a clean null at a wrong figure. The error was
not random; it was a fixed offset between the measurement
instrument's effective reference and the design reference. Type B
information that should have characterised this offset (the
manufacturing certificate for the metering rod, the assembly
record showing rod orientation) was either not propagated or not
incorporated into the figure uncertainty.

The mirror system was tested by two additional instruments before
launch: a refractive null corrector that used a different optical
principle, and an inverse null corrector that placed a calibrated
test surface in the optical path so the assembly could be checked
against a known reference. Both instruments indicated discrepancies
from the RNC reading. The refractive corrector's reading
differed by an amount consistent with the eventual on-orbit
figure error.

\paragraph{Organisational context.}
The optical metrology was performed at the contractor's plant
under cost and schedule pressure during the early-1980s
fabrication. The contract paid for completion to the RNC's null
and gave no incentive to invest additional time in cross-check
investigations. When the auxiliary instruments disagreed with the
RNC, the responsible engineering organisation classified the
auxiliary readings as instrument problems (the refractive
corrector was deemed less stable; the inverse corrector's
calibration was questioned) and treated the RNC as authoritative.
No analysis was conducted that would have used the disagreement
itself as a measurement signal: a quantity that three instruments
each measure with different mechanisms but agree on closely is
well-known; a quantity that one instrument reads with a small
stated uncertainty but two others disagree with by more than the
stated uncertainty is under-characterised. The reported uncertainty
on the mirror figure relied on the RNC alone.

Subsequent OSBI testimony established that some of the cross-check
disagreement was visible during fabrication but did not propagate
to the figure quality decision. The Board explicitly identified
the absence of an independent end-to-end optical test of the
mirror as a precipitating condition; a single hour on a flat-
mirror or autocollimation test against an external reference
would have revealed the error before launch
\cite{acc:nasa-hubble-optical-systems-1990}.

\paragraph{Lesson for uncertainty.}
A measurement system whose primary instrument reads a quantity to
small stated uncertainty, while two independent cross-check
instruments disagree by more than that uncertainty, is not a
measurement system reporting honestly. The engineering response
to such disagreement is one of three: (i) repeat each measurement
to rule out single-instrument artefact; (ii) identify a third
independent measurement that breaks the tie; (iii) widen the
reported uncertainty to a band that encompasses all three
readings until the cause of disagreement is identified and
corrected. The Hubble process took none of these paths; it
treated the preferred instrument as authoritative and the
disagreement as noise.

The lesson generalises beyond optics. Whenever an uncertainty
budget rests on a single dominant instrument and other
instruments disagree, the discipline is to widen the band, not
to narrow the source. The cost of recovering Hubble's design
performance through the 1993 COSTAR servicing mission was
approximately \$700 million; an additional optical cross-check
during fabrication would have cost a small fraction of that.

The Hubble case is recorded in the named-cases registry at
\repopath{docs/research/accidents/hubble-primary-mirror-1990.md}.

\subsection{The chronic class}

The same pattern appears in weaker forms whenever the reporting
process can benefit from a smaller interval: a test scenario
selected to preserve a passing result, a specification quoted
without its coverage basis, a sampling margin reported as if it
included non-response bias, a cost estimate collapsed from a range
to a single favorable point. One named case carries the pattern at
regulatory scale: Volkswagen Group's 2009--2015 light-duty diesel
vehicles were equipped with engine-control software that detected
the EPA's certified emissions test cycles and switched the
emissions-control system into a compliant calibration only during
testing; under ordinary driving the same vehicles emitted nitrogen
oxides at up to forty times the regulatory limit
\cite{acc:epa-vw-nov-2015, acc:doj-vw-2017}. Each example, named or
unnamed, requires the same audit question: what data and model
produced the interval?

The common failure is an uncertainty budget that answers to pressure
rather than evidence. Prevention is procedural: derive the
uncertainty from the data and model, document the Type A, Type B,
or Monte Carlo route, report the dominant component, and make the
calculation reproducible.

\subsection{The principle, restated}

\begin{principle}[Uncertainty derives from data, not preference]
The uncertainty of a measurement is computed from the data, the
instruments, the procedures, and the propagation of input
uncertainties through the measurement model. The uncertainty is
not chosen, negotiated, or selected for institutional convenience.
A reported uncertainty without a documented derivation is an
incomplete measurement result; the engineering action is to request
the uncertainty budget before using the number in a decision.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Building height by three methods, with error bars]
The reader measures the height of a chosen building by three
independent methods, computes the uncertainty for each, and
compares the three results to a published reference height. The
project tests the chapter's main mechanism: independent measurements
with documented uncertainty propagation should agree within their
combined error bars.

\textbf{Track}. Household and analysis (hybrid). The reader spends
time outdoors at street level near the building; no climbing, no
working at height, no instruments above battery voltage.
\textbf{Hazard class}: low.
\textbf{Tools}. Smartphone (clinometer or theodolite app, compass,
GPS, camera with EXIF or known focal length), tape measure
(preferably $\geq 10 \,\si{\meter}$), a stopwatch, a meter rule,
and a published reference height for the building (Wikipedia,
OpenStreetMap, building permit records, building owner's
documentation).

\textbf{Procedure}. The reader chooses a building whose published
height is well documented and whose base is accessible from a
distance of at least one building-height. They then perform three
of the following methods, computing each result and its standard
uncertainty using the linear propagation formula from section 4.3.

\begin{enumerate}[leftmargin=*]
  \item \engterm{Trigonometric method}. Measure a baseline distance
    $d$ from the reader's position to the base of the building.
    Measure the elevation angle $\theta$ from the reader's eye level
    to the top of the building using a smartphone clinometer.
    Compute height $h = d \tan\theta + h_{\text{eye}}$, where
    $h_{\text{eye}}$ is the reader's eye height above the ground.
    Propagate uncertainties from $u(d)$, $u(\theta)$, and
    $u(h_{\text{eye}})$ to $u(h)$ using the algebraic rules of
    section 4.3 applied to the small-angle approximation
    $\tan\theta \approx \theta$ near working angles, or by the
    finite-difference method (compute $h$ for nominal inputs and for
    inputs perturbed by their standard uncertainties; the spread is
    the propagated uncertainty estimate).
  \item \engterm{Shadow method}. On a sunny day, measure the
    building's shadow length $L_{\text{b}}$ and the reader's shadow
    length $L_{\text{r}}$ at the same time. Use the proportionality
    $h_{\text{building}} / h_{\text{reader}} = L_{\text{b}} /
    L_{\text{r}}$. Propagate uncertainties from $u(L_{\text{b}})$,
    $u(L_{\text{r}})$, and $u(h_{\text{reader}})$ to
    $u(h_{\text{building}})$.
  \item \engterm{Photographic method}. From a known distance $d$,
    take a photograph of the building with a known focal length and
    a known mapping from pixels to sensor length. Convert the
    measured pixel height to a sensor-plane height $s_{\text{image}}$
    in the same length units as the focal length $f$. For small
    angles and a camera aimed approximately perpendicular to the
    building face, use $h \approx d \cdot s_{\text{image}} / f$;
    otherwise compute the angular height from the camera geometry
    and use the trigonometric method. Propagate uncertainties from
    $d$, pixel measurement, focal length, sensor scaling, and camera
    alignment.
  \item \engterm{Hand-rule method}. First calibrate the reader's own
    hand rule: measure arm length and the width of the fist or
    thumb to compute its angular width. Stand at a safe measured
    distance where the building's apparent angular size can be
    compared with that calibrated angle. Compute the height from the
    angle and distance, and propagate the larger angular uncertainty
    explicitly.
\end{enumerate}

\textbf{Record}. Each method's record includes: measured input
quantities with their standard uncertainties, the computed height,
the propagated standard uncertainty, the expanded uncertainty at
$k = 2$, and the dominant uncertainty source identified by
sensitivity analysis (which input's exponent or coefficient amplifies
its relative uncertainty most strongly into the result).

\textbf{Analysis}. The reader produces a small comparison table
giving each method's result with $k = 2$ expanded uncertainty,
the published reference height with its source, and a column
showing whether the reference falls within each method's expanded
uncertainty range. Where the methods disagree by more than their
combined expanded uncertainty, the reader investigates why.

\textbf{Reflection (1500 words)}. The reader writes a 1500-word
narrative addressing:

\begin{itemize}[leftmargin=*]
  \item Which method gave the smallest error bar and why.
  \item Which method's dominant uncertainty source surprised the
    reader.
  \item Where two methods disagreed beyond their combined
    uncertainty, what mechanism the reader identifies as the most
    likely cause of the disagreement.
  \item Whether the published reference height fell within all
    three methods' expanded uncertainty, and what to do if it
    did not.
  \item One measurement in the reader's life or work where the
    same three-method discipline would change the reader's
    confidence in the result.
\end{itemize}

\textbf{Deliverable}. The three method records, the comparison
table, and the reflection. The general editor's reference set of
three-method computations for a representative building is
published online; the reader is encouraged to compare \emph{after}
completing their own.

\textbf{Time}. One to two hours per method outdoors, plus two to
three hours of analysis and writing. Total approximately six to
nine hours, distributed across two or three days.
\end{project}

\section{Exercises}

The exercises train calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment. Calculation and
derivation problems have full solutions; open-ended problems have
rubrics or reference write-ups. Exercises are tagged with their
reader-path tier.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A digital voltmeter reads $5.32 \,\si{\volt}$ with a manufacturer-
specified accuracy of $\pm (0.05\,\% \text{ of reading} + 1 \text{
last digit})$. The last digit is at the $10 \,\si{\milli\volt}$
position. Compute the standard uncertainty assuming the
specification is at $k = 2$ and the underlying distributions are
Gaussian.
\end{exercise}

\paragraph{Solution sketch.}
The reading-proportional term is $0.0005 \times 5.32 \,\si{\volt} =
2.66 \,\si{\milli\volt}$. The last-digit term is
$10 \,\si{\milli\volt}$. Summed as expanded uncertainties at $k = 2$
the total is $12.66 \,\si{\milli\volt}$. Dividing by $k = 2$ gives
the standard uncertainty $u \approx 6.3 \,\si{\milli\volt}$. (A
stricter analysis treats the two contributions as independent and
combines them in quadrature at the standard level: $u =
\sqrt{(2.66/2)^{2} + (10/2)^{2}} \approx 5.2 \,\si{\milli\volt}$.
Report the convention used.)

\begin{exercise}
Two thermometers read the same room temperature: thermometer A
reads $21.4 \,\si{\degreeCelsius}$ with $u_{A} = 0.2 \,\si{\degreeCelsius}$,
thermometer B reads $21.7 \,\si{\degreeCelsius}$ with $u_{B} = 0.3
\,\si{\degreeCelsius}$. Compute the weighted mean and its standard
uncertainty, assuming the two readings are independent.
\end{exercise}

\paragraph{Solution sketch.}
Weight by inverse variance: $w_{A} = 1/0.04 = 25$,
$w_{B} = 1/0.09 \approx 11.1$. The weighted mean is
$(25 \times 21.4 + 11.1 \times 21.7)/(25 + 11.1) = (535.0 + 240.9)/
36.1 \approx 21.50 \,\si{\degreeCelsius}$. The standard uncertainty
of the weighted mean is $1/\sqrt{w_{A} + w_{B}} = 1/\sqrt{36.1}
\approx 0.17 \,\si{\degreeCelsius}$. Note that $u_{w}$ is smaller
than either input's uncertainty, as expected when independent
information is combined.

\begin{exercise}
A length is measured as $L = 1.250 \,\si{\meter}$ with a maximum-
error specification of $\pm 0.002 \,\si{\meter}$ and no stated
distribution. Treat the error as uniformly distributed on
$[-0.002, 0.002] \,\si{\meter}$ and compute the standard uncertainty.
\end{exercise}

\paragraph{Solution sketch.}
For a uniform distribution on $[-a, +a]$, the standard deviation is
$a/\sqrt{3}$. With $a = 0.002 \,\si{\meter}$, the standard
uncertainty is $u(L) = 0.002/\sqrt{3} \approx 0.00115 \,\si{\meter}
= 1.15 \,\si{\milli\meter}$. The naive treatment that takes
$0.002 \,\si{\meter}$ as a standard uncertainty overstates the
uncertainty by a factor of $\sqrt{3} \approx 1.73$; the naive
treatment that takes $0.002 \,\si{\meter}$ as the half-width of a
$k = 2$ Gaussian expanded uncertainty (so $u = 0.001 \,\si{\meter}$)
understates by a small but non-negligible amount.

\begin{exercise}
A rectangle has measured side lengths $a = 12.5 \,\si{\centi\meter}
\pm 0.1 \,\si{\centi\meter}$ and $b = 8.3 \,\si{\centi\meter} \pm
0.1 \,\si{\centi\meter}$ (both at $k = 1$). Compute the area and
its standard uncertainty.
\end{exercise}

\paragraph{Solution sketch.}
The area is $A = ab = 12.5 \times 8.3 = 103.75 \,\si{\centi\meter
\squared}$. The relative uncertainties are $u(a)/a = 0.008$ and
$u(b)/b = 0.012$. By the product rule, $(u(A)/A)^{2} =
0.008^{2} + 0.012^{2} = 6.4 \times 10^{-5} + 1.44 \times 10^{-4}
= 2.08 \times 10^{-4}$, so $u(A)/A \approx 0.0144$ and $u(A)
\approx 1.5 \,\si{\centi\meter\squared}$. Report
$A = 104 \pm 2 \,\si{\centi\meter\squared}$ (rounded with the
uncertainty determining the last significant figure).

\begin{exercise}
A current $I = 2.0 \,\si{\ampere} \pm 0.05 \,\si{\ampere}$ flows
through a resistor $R = 100 \,\si{\ohm} \pm 1 \,\si{\ohm}$. Compute
the dissipated power and its standard uncertainty, assuming the
two measurements are uncorrelated. Identify which input contributes
more to the power's uncertainty.
\end{exercise}

\paragraph{Solution sketch.}
$P = I^{2}R = 4 \times 100 = 400 \,\si{\watt}$. By the
product-of-powers rule with exponents $+2$ on $I$ and $+1$ on $R$:
$(u(P)/P)^{2} = 4 (u(I)/I)^{2} + 1 (u(R)/R)^{2} =
4 \times 0.025^{2} + 0.01^{2} = 0.0025 + 0.0001 = 0.0026$. So
$u(P)/P \approx 0.051$ and $u(P) \approx 20 \,\si{\watt}$. The
current dominates: its relative-uncertainty contribution
$0.0025$ is twenty-five times the resistance's $0.0001$, because
the exponent-squared factor of $4$ amplifies $I$'s already-larger
relative uncertainty.

\begin{exercise}
A measurement has a value of $9.876 \,\si{\meter\per\second\squared}$
with a standard uncertainty of $0.043 \,\si{\meter\per\second
\squared}$. Report the value at coverage factor $k = 2$ in (a) the
$\pm$ form, (b) the parenthesis form, with both rounded to
appropriate decimal places.
\end{exercise}

\paragraph{Solution sketch.}
Expanded uncertainty: $U = k \cdot u = 2 \times 0.043 = 0.086 \,\si
{\meter\per\second\squared}$. Round the uncertainty to two
significant figures because the leading digit is small (the rule
of section 4.5.2): $U = 0.09 \,\si{\meter\per\second\squared}$.
Round the value to the same decimal place. (a) $\pm$ form:
$g = 9.88 \pm 0.09 \,\si{\meter\per\second\squared} \;(k = 2)$.
(b) Parenthesis form: $g = 9.88(9) \,\si{\meter\per\second\squared}
\;(k = 2)$. State the coverage factor in either case.

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Starting from the algebraic sum rule, derive the combined standard
uncertainty for $y = a x_{1} + b x_{2} + c x_{3}$ with all inputs
uncorrelated. State the conditions under which the derivation is
valid (linearity, independence, finite variance) and identify the
case in which the formula understates the combined uncertainty.
\end{exercise}

\paragraph{Solution.}
The two-input sum rule, derived in the chapter, states that for
$w = a x_{1} + b x_{2}$ with $x_{1}$ and $x_{2}$ uncorrelated,
$u^{2}(w) = a^{2} u^{2}(x_{1}) + b^{2} u^{2}(x_{2})$. The three-input
case extends by treating $y$ as the sum of $w$ and a third term.

\emph{Step 1: pair the first two inputs.} Let $w = a x_{1} + b x_{2}$.
By the two-input sum rule,
\[
u^{2}(w) = a^{2} u^{2}(x_{1}) + b^{2} u^{2}(x_{2}).
\]

\emph{Step 2: combine $w$ with the third input.} We need to show
that $w$ and $x_{3}$ are uncorrelated. Since $w$ is a linear
combination of $x_{1}$ and $x_{2}$ alone, $\Cov(w, x_{3}) =
a\Cov(x_{1}, x_{3}) + b\Cov(x_{2}, x_{3})$. Under the assumption
that all three inputs are pairwise uncorrelated, both covariances
vanish, so $\Cov(w, x_{3}) = 0$. The pair-and-conquer construction
is sound.

\emph{Step 3: apply the two-input sum rule again.} Write
$y = w + c x_{3}$ with $w$ and $x_{3}$ uncorrelated:
\[
u^{2}(y) = u^{2}(w) + c^{2} u^{2}(x_{3})
= a^{2} u^{2}(x_{1}) + b^{2} u^{2}(x_{2}) + c^{2} u^{2}(x_{3}).
\]
This is the claimed result.

\emph{Step 4: validity conditions.} Three conditions enter the
derivation: (i) $y$ is linear in the inputs (the rule is exact here
because $y$ is a linear combination, with no higher-order Taylor
terms); (ii) the three inputs are pairwise independent (no
covariance terms appear); (iii) each input has finite variance
(otherwise $u^{2}(x_{i})$ is undefined and the formula fails to
make sense, as the Cauchy case illustrates).

\emph{Step 5: the correlated case.} If $x_{1}$ and $x_{2}$ have
correlation coefficient $\rho_{12}$, the two-input sum rule picks
up a covariance term:
\[
u^{2}(w) = a^{2} u^{2}(x_{1}) + b^{2} u^{2}(x_{2}) + 2ab\rho_{12}
u(x_{1}) u(x_{2}).
\]
Tracking this through the three-input combination and allowing the
remaining pairs $(x_{1}, x_{3})$ and $(x_{2}, x_{3})$ to carry
their own correlation coefficients gives the general formula
\[
u^{2}(y) = a^{2} u^{2}(x_{1}) + b^{2} u^{2}(x_{2}) + c^{2}
u^{2}(x_{3}) + 2ab\rho_{12}u(x_{1})u(x_{2}) + 2ac\rho_{13}u(x_{1})
u(x_{3}) + 2bc\rho_{23}u(x_{2})u(x_{3}).
\]
The uncorrelated formula understates the combined uncertainty when
any of the products $\rho_{ij}u(x_{i})u(x_{j})$ is positive with a
matching coefficient sign. The understatement can be large: at
$\rho = 1$ and equal $u$, the missing cross-term is the same size
as each of the diagonal terms it accompanies.

\emph{Step 6: the engineering point.} A measurement chain in which
three terms share a common reference (the same calibration
standard, the same environmental sensor, the same operator
technique) has positive correlations among the corresponding
uncertainty contributions. The uncorrelated formula is the GUM's
working default; a serious uncertainty budget identifies the
correlations explicitly and either adds the covariance terms or
documents why the correlation can be neglected for the budget at
hand. The default is conservative when the correlations are
negative and optimistic when they are positive; only the budget
that names the correlation discipline can be audited.

\begin{exercise}
Show that, in the small-error limit, the relative uncertainty of a
product $z = xy$ satisfies
\[
\left(\frac{u(z)}{z}\right)^{2}
= \left(\frac{u(x)}{x}\right)^{2}
+ \left(\frac{u(y)}{y}\right)^{2}
\]
when $x$ and $y$ are uncorrelated. Identify a domain in which this
formula systematically underestimates the uncertainty.
\end{exercise}

\paragraph{Solution.}
The relative-uncertainty formula for products is the multiplicative
counterpart of the absolute-uncertainty sum rule. Working in
relative variables makes the derivation transparent.

\emph{Step 1: write each input in relative form.} Let $x =
x_{0}(1 + \epsilon_{x})$ and $y = y_{0}(1 + \epsilon_{y})$ where
$x_{0}, y_{0}$ are the nominal values and $\epsilon_{x},
\epsilon_{y}$ are the relative deviations (small fractions). The
nominal product is $z_{0} = x_{0} y_{0}$.

\emph{Step 2: expand the product.} The product $z = xy$ is
\[
z = x_{0} y_{0}(1 + \epsilon_{x})(1 + \epsilon_{y}) = z_{0}(1 +
\epsilon_{x} + \epsilon_{y} + \epsilon_{x}\epsilon_{y}).
\]
Define the relative deviation of $z$ as $\epsilon_{z} = (z - z_{0})/
z_{0} = \epsilon_{x} + \epsilon_{y} + \epsilon_{x}\epsilon_{y}$.

\emph{Step 3: drop the second-order term.} For small relative
uncertainties, $|\epsilon_{x}|, |\epsilon_{y}| \ll 1$, the product
$\epsilon_{x}\epsilon_{y}$ is much smaller than the linear terms
and can be dropped. This gives $\epsilon_{z} \approx \epsilon_{x} +
\epsilon_{y}$.

\emph{Step 4: compute the variance of the relative deviation.} If
$\epsilon_{x}$ and $\epsilon_{y}$ are uncorrelated random
variables, then
\[
u^{2}(\epsilon_{z}) = u^{2}(\epsilon_{x}) + u^{2}(\epsilon_{y}),
\]
and since $\epsilon = u(\cdot)/(\cdot)_{0}$ for each variable,
this is
\[
\left(\frac{u(z)}{z_{0}}\right)^{2} = \left(\frac{u(x)}{x_{0}}
\right)^{2} + \left(\frac{u(y)}{y_{0}}\right)^{2},
\]
the relative-uncertainty product rule.

\emph{Step 5: identify the underestimating domain.} The derivation
underestimates $u(z)$ in three scenarios. \emph{Large relative
uncertainties}: the dropped $\epsilon_{x}\epsilon_{y}$ term is of
order $u(x)/x \cdot u(y)/y$, which for $u/x = u/y = 0.1$ is
$0.01$, a $10\%$ contribution to the linear term; the rule
underestimates by that amount. \emph{Positive correlation between
$x$ and $y$}: the missing covariance $\Cov(\epsilon_{x},
\epsilon_{y}) = \rho u(\epsilon_{x}) u(\epsilon_{y})$ adds to the
sum, and the rule with $\rho > 0$ missing understates $u(z)$.
\emph{Heavy-tailed input distributions}: when the variance is not
a good summary of the spread (Cauchy, Pareto with shape
$\alpha \leq 2$), the quadrature combination is itself
misleading; the rule is not so much underestimating as inapplicable.

\emph{Step 6: a check by the partial-derivative form.} The general
GUM rule for $z = f(x, y)$ with uncorrelated inputs is $u^{2}(z) =
(\partial f/\partial x)^{2} u^{2}(x) + (\partial f/\partial y)^{2}
u^{2}(y)$. For $z = xy$, $\partial z/\partial x = y$ and
$\partial z/\partial y = x$, giving $u^{2}(z) = y_{0}^{2} u^{2}(x)
+ x_{0}^{2} u^{2}(y)$. Dividing by $z_{0}^{2} = x_{0}^{2} y_{0}^{2}$
recovers the relative-uncertainty form. The relative formula and
the partial-derivative formula are the same statement.

\begin{exercise}
Derive the propagation rule for $z = x + y$ when $x$ and $y$ have
correlation coefficient $\rho$. State the result, and identify the
case in which the uncertainty of $z$ is smaller than either $u(x)$
or $u(y)$.
\end{exercise}

\paragraph{Solution sketch.}
$\Var(z) = \Var(x) + \Var(y) + 2 \Cov(x, y) = u^{2}(x) + u^{2}(y)
+ 2 \rho u(x) u(y)$. The sum-rule special case is
$\rho = 0$. For $z = x - y$ the cross-term flips sign. Now consider
when $u(z) < \min(u(x), u(y))$: this never happens for $z = x + y$
because each variance term and the cross-term (when $\rho \geq 0$)
are non-negative. For $z = x - y$ with $\rho = +1$ and
$u(x) = u(y)$, however, $u(z) = 0$: the correlated component
cancels completely. The exercise question is a probe of the sign
convention; in the sum case, no choice of $\rho$ reduces $u(z)$
below the larger of $u(x), u(y)$ unless $\rho < 0$ and the inputs
are of similar magnitude.

\begin{exercise}
For a quantity $y = f(x)$ with $x$ uniformly distributed on
$[a, b]$, derive the standard uncertainty of $y$ assuming $f$ is
linear over the interval. Compare to the result for $x$ Gaussian
with the same standard deviation.
\end{exercise}

\paragraph{Solution.}
The first-order propagation rule treats only the variance of the
input, not its distribution. The solution makes that explicit and
then identifies where the distribution does enter the working
calculation.

\emph{Step 1: write down the variance of $y$.} For
$f(x) = c_{0} + c_{1} x$ the variance scales as $\Var(y) =
c_{1}^{2} \Var(x)$ for any distribution of $x$, since variance is
preserved under linear transformations up to the squared
multiplier. The shift $c_{0}$ does not affect variance.

\emph{Step 2: compute the standard deviation for the uniform input.}
A uniform distribution on $[a, b]$ has mean $(a + b)/2$ and
variance $(b - a)^{2}/12$ (the standard result from probability
theory; the second-moment integral $\int_{a}^{b}(x - \mu)^{2}/(b-a)
\,\dd x$ evaluates to this). Therefore
\[
u(y) = |c_{1}| \sqrt{(b - a)^{2}/12} = |c_{1}| (b-a)/(2\sqrt{3}).
\]

\emph{Step 3: compute the standard deviation for the Gaussian
input.} A Gaussian $x$ with standard deviation $\sigma_{x} =
(b-a)/(2\sqrt{3})$ produces $u(y) = |c_{1}| \sigma_{x} =
|c_{1}|(b-a)/(2\sqrt{3})$, identical to the uniform result by
construction. The first-order propagation rule gives the same
standard uncertainty for the two inputs once they share the same
variance.

\emph{Step 4: identify what the shape determines.} The standard
uncertainty is the same, but the distribution shape of $y$ differs.
Uniform input under linear $f$ gives uniform $y$ output; Gaussian
input gives Gaussian output. The coverage intervals at fixed
probability differ:
\begin{itemize}
  \item Gaussian: the $95\%$ interval is $\pm 1.96\sigma_{y}$,
    the standard normal critical value times the standard deviation.
  \item Uniform: the $95\%$ interval is $\pm 0.95 \cdot \sqrt{3}
    \sigma_{y} \approx \pm 1.65 \sigma_{y}$, the $95\%$ half-width
    of a uniform distribution scaled to the same standard
    deviation.
\end{itemize}
The two intervals differ by about $20\%$ at the $95\%$ coverage
level.

\emph{Step 5: the engineering point.} The standard uncertainty is
distribution-blind under first-order propagation. The expanded
uncertainty at a stated coverage probability is not: it depends on
the output distribution, which depends on the input distribution
and the propagation model. A measurement budget that reports the
standard uncertainty and the coverage factor implicitly assumes a
distribution; the report should state which one. The GUM's default
of $k = 2$ corresponds to Gaussian outputs at $95\%$ coverage; for
other distributions the same $k$ corresponds to a different
coverage probability.

\emph{Step 6: a generalisation.} The same logic applies to any
non-Gaussian input: triangular ($\sigma^{2} = a^{2}/6$ for
half-width $a$), trapezoidal, U-shaped (resonant systems). The
standard-deviation calculation is geometry-specific; the
propagation through a linear model is geometry-blind; the
coverage interval at a stated probability is again
geometry-specific. A serious budget separates the three.

\begin{exercise}
Show that the standard uncertainty of the mean of $n$ independent
measurements, each with standard uncertainty $\sigma$, is
$\sigma / \sqrt{n}$. Then consider the simpler correlated case in
which every pair of measurements has the same correlation
coefficient $\rho$. Derive the standard uncertainty of the mean and
state why positive correlation reduces the benefit of averaging.
\end{exercise}

\paragraph{Solution sketch.}
For independent inputs, $\bar{x} = (1/n)\sum x_{i}$ has variance
$\Var(\bar{x}) = (1/n^{2})\sum \Var(x_{i}) = \sigma^{2}/n$, so
$u(\bar{x}) = \sigma/\sqrt{n}$. For equicorrelated inputs with
common correlation $\rho$, $\Var(\bar{x}) = (1/n^{2})[n\sigma^{2}
+ n(n-1)\rho\sigma^{2}] = (\sigma^{2}/n)[1 + (n-1)\rho]$. As
$n \to \infty$ the standard uncertainty of the mean approaches
$\sigma\sqrt{\rho}$ rather than zero. Even a small positive $\rho$
sets a floor: $\rho = 0.01$ gives a long-$n$ limit of
$0.1\sigma$, which means averaging beyond a few hundred
measurements yields no further reduction. The mechanism is that
correlation injects a shared common-mode error that all readings
record together, and averaging cannot suppress what every reading
records identically.

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the standard uncertainty of a smartphone clinometer's
elevation angle reading using assumed values supplied in the
problem: bias of order $10^{-3} \,\si{\radian}$ after calibration
and short-term noise of order $10^{-4} \,\si{\radian}$. State
whether your result is controlled by bias or random noise, and name
the measurement you would perform to replace these assumed values
with data.
\end{exercise}

\paragraph{Solution.}
The clinometer estimate combines a Type B bias contribution with a
Type A random contribution. The solution computes the combined
standard uncertainty, identifies the dominant component, and
specifies the experiment that replaces assumed values with measured
ones.

\emph{Step 1: identify the two contributions.} The problem states
a calibration bias of $u_{B} \approx 10^{-3}\,\si{\radian}$ and a
short-term noise contribution of $u_{A} \approx 10^{-4}\,\si{
\radian}$. Both are standard uncertainties at $k = 1$.

\emph{Step 2: combine in quadrature.} For independent
contributions the combined standard uncertainty is
\[
u(\theta) = \sqrt{u_{B}^{2} + u_{A}^{2}} = \sqrt{(10^{-3})^{2} +
(10^{-4})^{2}} \approx 1.005 \times 10^{-3}\,\si{\radian} \approx
10^{-3}\,\si{\radian}.
\]
The combined uncertainty is essentially equal to the larger
contribution alone. In angular degrees, $10^{-3}\,\si{\radian}
\approx 0.057^{\circ}$, or about $3.4$ arcminutes.

\emph{Step 3: identify the dominant component.} The bias-to-random
ratio is $u_{B}/u_{A} = 10$, so $u_{B}^{2}$ exceeds $u_{A}^{2}$ by
a factor of $100$. The bias dominates the combined uncertainty by
two orders of magnitude in the variance budget.

\emph{Step 4: confirm averaging cannot reduce the dominant
component.} Averaging $n$ independent readings reduces the random
contribution by $\sqrt{n}$ but leaves the systematic component
unchanged. At $n = 100$, the random standard uncertainty falls to
$10^{-4}/\sqrt{100} = 10^{-5}\,\si{\radian}$, while the bias
remains $10^{-3}\,\si{\radian}$. The combined uncertainty becomes
$\sqrt{(10^{-3})^{2} + (10^{-5})^{2}} \approx 10^{-3}\,\si{\radian}$,
no improvement. Averaging more readings hits the systematic floor
quickly and yields nothing further.

\emph{Step 5: design the experiment that replaces assumed values.}
The bias estimate requires a comparison against a known reference.
Place the phone against a precision surface plate or a calibrated
wedge at known angles (the conventional set is
$0^{\circ}, 30^{\circ}, 45^{\circ}, 60^{\circ}, 90^{\circ}$) and
record the clinometer indication at each. Repeat the measurement
at each angle multiple times. From the data,
\begin{itemize}[leftmargin=*]
  \item the mean deviation across the calibrated angles is the
    bias estimate (replacing the assumed $10^{-3}$);
  \item the standard deviation across repeated readings at one
    angle is the random component estimate (replacing the assumed
    $10^{-4}$);
  \item the spread of bias estimates across the angle range
    detects angle-dependent biases (a single calibration offset
    is one possibility; a sinusoidal angle dependence from a
    sensor misalignment is another, with very different mitigation).
\end{itemize}

\emph{Step 6: the engineering point.} A smartphone clinometer's
typical bias is the dominant uncertainty source and cannot be
averaged out. The right engineering action is calibration against
a known reference, not better averaging. The chapter's discipline
of separating Type A from Type B contributions surfaces this
distinction: if the operator believes averaging will improve a
measurement, they have implicitly assumed the bias is small, and
the chapter's compatibility-ratio principle is the test of that
assumption against an independent reference.

\begin{exercise}
Estimate how many independent voltage measurements would be required
to reduce a standard uncertainty of $10 \,\si{\milli\volt}$ to
$1 \,\si{\milli\volt}$, assuming the dominant uncertainty is
random with no systematic component. Then estimate the same
required count when half the uncertainty is systematic.
\end{exercise}

\paragraph{Solution sketch.}
Pure-random case: $\sigma/\sqrt{n} = 1$, so
$n = (\sigma/u_{\text{target}})^{2} = (10/1)^{2} = 100$
measurements. Half-systematic case: split as
$u^{2} = u_{r}^{2} + u_{s}^{2}$ with $u_{r}^{2} = u_{s}^{2} =
50 \,\si{\milli\volt\squared}$, so $u_{r} = u_{s} = \sqrt{50}
\approx 7.07 \,\si{\milli\volt}$. The systematic component
$u_{s}$ does not average out. The combined uncertainty after
averaging $n$ readings is $\sqrt{50/n + 50}$. To reach
$1 \,\si{\milli\volt}$ requires $50/n + 50 \leq 1$, which is
impossible (the systematic floor alone is seven times the
target). The lesson: systematic-uncertainty reduction requires
recalibration or method change, not more averaging.

\begin{exercise}
Estimate the standard uncertainty of the mean height of a
representative adult population from a sample of $30$ measurements
each with standard uncertainty $0.5 \,\si{\centi\meter}$. State
the assumptions about between-individual variability versus
within-individual measurement uncertainty.
\end{exercise}

\paragraph{Solution.}
The exercise distinguishes the measurement uncertainty on a single
height reading from the population-level variability in adult
height. Both contribute to the standard uncertainty of the
sample-mean estimate, but the second dominates.

\emph{Step 1: identify the two variability sources.} A single
measurement of a single person's height has two independent
random contributions: (i) the instrument's measurement uncertainty
($0.5\,\si{\centi\meter}$, given) and (ii) the variation between
individuals in the population, which for a representative adult
sample has standard deviation around $7\,\si{\centi\meter}$ (typical
working value for a single national or regional adult population).

\emph{Step 2: combine the per-measurement contributions.} The two
contributions are independent, so the per-measurement standard
deviation is
\[
\sigma_{1} = \sqrt{u_{\text{meas}}^{2} + \sigma_{\text{pop}}^{2}} =
\sqrt{0.5^{2} + 7^{2}} = \sqrt{49.25} \approx 7.0\,\si{\centi
\meter}.
\]
The measurement contribution is negligible against the
population's spread.

\emph{Step 3: compute the standard uncertainty of the sample
mean.} For $n = 30$ independent measurements,
\[
u(\bar{x}) = \sigma_{1}/\sqrt{n} = 7.0/\sqrt{30} \approx 1.28\,
\si{\centi\meter}.
\]
The reported sample-mean uncertainty is about $1.3\,\si{\centi
\meter}$ at $k = 1$.

\emph{Step 4: distinguish sample mean from population mean.} The
$\bar{x} \pm 1.3\,\si{\centi\meter}$ estimate is the uncertainty
of the mean of the \emph{sampled} group. Whether $\bar{x}$
estimates the \emph{population} mean depends on the sample's
representativeness. If the $30$ people happen to be selected from
a basketball team, the sample mean reflects that subpopulation,
not the general adult population, regardless of the per-measurement
precision.

\emph{Step 5: identify the floor that more measurements cannot
reduce.} The $1/\sqrt{n}$ scaling of $u(\bar{x})$ assumes the
sample is a random draw from the population. Sampling bias (who
shows up to be measured) sets a floor that more measurements
cannot reduce. A street-corner survey in a particular
neighbourhood may have $u(\bar{x}) < 1\,\si{\centi\meter}$ at
$n = 200$ but estimate the wrong population mean by $5\,\si{\centi
\meter}$. Increasing $n$ reduces the statistical uncertainty
without addressing the bias.

\emph{Step 6: the engineering point.} The chapter's discipline
applies: the measurement uncertainty contributes negligibly
compared to the natural population variability. The sample-mean
uncertainty calculation is straightforward; the bias question
(does the sample represent the population?) is the harder one.
For a working population-mean estimate, the sample design (random
selection across the population) matters more than the per-
measurement precision. The $\bar{x} \pm 1.3\,\si{\centi\meter}$
result is correct as a statement about the sample; extending it
to the population requires the sampling-bias question be answered
separately.

\begin{exercise}
Estimate the height of the tallest building you can see from the
window of your home or workplace, with an explicit error bar
derived from your own assumptions about distance and angular
size. Compare your error bar to the published reference height.
\end{exercise}

\paragraph{Solution sketch.}
Example calculation for a building $200 \,\si{\meter}$ away
subtending $\theta \approx 25^{\circ} = 0.44 \,\si{\radian}$ at
the eye. With $u(d) \approx 20 \,\si{\meter}$ (10\% from
map-eyeballing) and $u(\theta) \approx 0.03 \,\si{\radian}$ (6\%
from hand-rule estimation), the height $h = d \tan\theta$ gives
$h \approx 93 \,\si{\meter}$ with relative uncertainty
$\sqrt{0.10^{2} + 0.06^{2}} \approx 0.12$, so $u(h) \approx
11 \,\si{\meter}$. The reader's specific numbers will differ; the
discipline is to declare both contributions explicitly and propagate
them, not to report a single point estimate.

\begin{exercise}
Estimate the relative uncertainty in a stopwatch-timed measurement
of the period of a pendulum, given a stopwatch reaction time of
$\sim 0.2 \,\si{\second}$. State the conditions under which the
reaction-time uncertainty is the dominant component of the
period's uncertainty.
\end{exercise}

\paragraph{Solution sketch.}
A reaction-time uncertainty enters twice (start and stop), so
$u(T_{\text{measured}}) \approx \sqrt{2} \times 0.2 \approx
0.28 \,\si{\second}$. For a single-period measurement of a
$T \approx 2 \,\si{\second}$ pendulum, this gives relative
uncertainty $\sim 14\%$, dominated by reaction time. To reduce it,
time $N$ periods at once and divide: timing $20$ periods gives
$T_{\text{total}} = 40 \,\si{\second}$ with the same
$0.28 \,\si{\second}$ uncertainty, for a per-period uncertainty
of $0.014 \,\si{\second}$ or $0.7\%$. Reaction time dominates
when (i) only one or a few periods are timed and (ii) no
electronic gate (photogate, smartphone microphone trigger)
replaces the human start-stop.

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
In a language of the reader's choice, simulate $10^{4}$ independent
samples of $z = x + y$ where $x \sim \mathcal{N}(0, 1)$ and
$y \sim \mathcal{N}(0, 1)$ are independent. Compute the empirical
standard deviation of $z$ and compare to the analytical result
$\sqrt{2}$. Then repeat with $\rho = 0.5$ and $\rho = -0.5$
correlation between $x$ and $y$.
\end{exercise}

\paragraph{Solution sketch.}
Independent case yields empirical $\sigma_{z} \approx 1.414$,
within about $1\%$ of $\sqrt{2}$ at $N = 10^{4}$. For $\rho =
0.5$ the analytical result is $\sqrt{1 + 1 + 2 \times 0.5}
= \sqrt{3} \approx 1.732$. For $\rho = -0.5$, $\sqrt{1 + 1 - 1}
= 1$. To simulate correlated normals, draw $(x, y)$ from a
bivariate normal with covariance matrix $\Sigma = [[1, \rho],
[\rho, 1]]$; in practice, generate $u, v$ independent
$\mathcal{N}(0,1)$ and form $x = u$, $y = \rho u + \sqrt{1 -
\rho^{2}} v$. Reference implementation in
\texttt{code/correlated\_sum.py}.

\begin{exercise}
Simulate $10^{4}$ independent measurements of a quantity drawn from
a uniform distribution on $[-1, 1]$. Estimate the empirical
distribution of the mean of $n$ such measurements for $n = 1, 5,
10, 50, 100$, and compare to the Gaussian prediction from the
central limit theorem. Identify the value of $n$ at which the
sample mean's distribution becomes visually indistinguishable from
Gaussian.
\end{exercise}

\paragraph{Solution sketch.}
The variance of the uniform distribution on $[-1, 1]$ is
$\sigma^{2} = 4/12 = 1/3$. The CLT predicts the mean of $n$
samples is approximately $\mathcal{N}(0, 1/(3n))$. The
flat-topped uniform shape transitions to bell-shaped quickly:
at $n = 5$ the histogram is already close to Gaussian; by
$n = 10$ the residual visible difference is in the tails (the
uniform is hard-bounded, so the mean cannot exceed $1$ even at
small $n$). For the well-behaved uniform parent, $n \approx 10$
suffices for visual Gaussianity. Reference implementation in
\texttt{code/clt\_uniform.py}.

\begin{exercise}
Implement a Monte Carlo propagation per JCGM 101 for the pendulum
$g$-measurement of section 4.3. Use Gaussian distributions for $L$
and $T$ with the standard deviations stated in the worked example.
Compute the empirical standard deviation of $g$ and compare to the
linear-propagation result $0.05 \,\si{\meter\per\second\squared}$.
Comment on whether the linear approximation is appropriate for this
problem.
\end{exercise}

\paragraph{Solution sketch.}
Draw $10^{5}$ samples of $L \sim \mathcal{N}(1.000, 0.001^{2})$
and $T \sim \mathcal{N}(2.007, 0.005^{2})$, compute
$g = 4\pi^{2} L/T^{2}$ for each, and take the empirical standard
deviation. The result is approximately
$0.049 \,\si{\meter\per\second\squared}$, within $2\%$ of the
linear-propagation prediction. The empirical distribution is
symmetric and Gaussian-shaped because the input
relative uncertainties are small (well below $1\%$) so the
linear approximation is excellent. Reference implementation in
\texttt{code/pendulum\_mc.py}; the comparison plot is in
\texttt{figures/fig-mc-vs-linear.tex}.

\begin{exercise}
Simulate the distribution of the sample mean of $n = 5$ samples
drawn from a Cauchy distribution. Plot the empirical distribution
of the sample mean for $1000$ trials and identify the failure of
the central limit theorem. Comment on what averaging more samples
would, and would not, accomplish.
\end{exercise}

\paragraph{Solution sketch.}
The sample mean of $n$ Cauchy variables is itself Cauchy with the
same scale parameter; averaging does not reduce the spread. The
empirical histogram shows a sharp peak near zero plus extreme
outliers reaching $\pm 100$ or more. Increasing $n$ to $1000$
makes no difference: the distribution of the mean remains Cauchy.
What averaging \emph{cannot} accomplish: variance reduction. What
it \emph{can} accomplish: a more reliable median (the median of
Cauchy samples does converge under sample-size increases) and a
sense of the parent distribution's scale. The takeaway is that
robust estimators (median, trimmed mean) replace the arithmetic
mean when the parent has heavy tails. Reference implementation in
\texttt{code/cauchy\_mean.py}.

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a measurement plan to determine the average mass of an
object in a population of $100$ nominally identical units with an
expanded uncertainty of $U = 0.5 \,\si{\gram}$ at $k = 2$. State
the required standard uncertainty of the mean, the instrument
accuracy or calibration uncertainty, the number of units sampled,
the number of measurements per sampled unit, and the procedure for
separating between-unit variability from measurement repeatability.
\end{exercise}

\paragraph{Discussion.}
Required $u(\bar{m}) = 0.25 \,\si{\gram}$. A balance with
calibration uncertainty $\leq 0.1 \,\si{\gram}$ is suitable;
calibrate against a known mass before and after the campaign and
record the drift. Sample at least $30$ units (a defensible
representative sample of $100$); measure each unit $3$ times in
random order. The mean of the per-unit means is the population
mean estimate. A one-way analysis of variance separates the
between-unit variance from the within-unit (repeatability)
variance. The standard uncertainty of the population mean is
$\sqrt{s_{\text{between}}^{2}/n_{\text{units}} +
s_{\text{within}}^{2}/(n_{\text{units}} \cdot n_{\text{rep}})
+ u_{\text{calibration}}^{2}}$. A rubric expects: target
$u(\bar{m})$, separate between/within decomposition, calibration
budget item, randomisation of measurement order.

\begin{exercise}
Design an uncertainty budget for a household refrigerator
temperature measurement with target expanded uncertainty
$U = 0.5 \,\si{\degreeCelsius}$ at $k = 2$. List each input source,
including thermometer calibration, sensor resolution, sampling
location, time-of-day variation, defrost cycle, and door opening,
with an estimated standard uncertainty contribution. Combine the
contributions and identify which component dominates.
\end{exercise}

\paragraph{Discussion.}
A representative budget at target $u = 0.25 \,\si{\degreeCelsius}$:
thermometer calibration $0.1 \,\si{\degreeCelsius}$
(certificate), sensor resolution $0.06 \,\si{\degreeCelsius}$
($0.1/\sqrt{3}$ for digital quantisation at $0.1 \,\si
{\degreeCelsius}$ resolution), spatial gradient inside the
refrigerator $0.5 \,\si{\degreeCelsius}$ (top shelf vs door
bin), time-of-day variation $0.3 \,\si{\degreeCelsius}$
(cooling-cycle oscillation), defrost transient
$0.4 \,\si{\degreeCelsius}$ (mitigate by sampling outside the
defrost window), door-opening events $0.5 \,\si{\degreeCelsius}$
(mitigate by sampling at consistent times). Combined in quadrature:
$\sqrt{0.1^{2} + 0.06^{2} + 0.5^{2} + 0.3^{2} + 0.4^{2} + 0.5^{2}}
\approx 0.9 \,\si{\degreeCelsius}$. Spatial gradient and door-
opening dominate. The target $u = 0.25 \,\si{\degreeCelsius}$ is
not achievable without specifying location and time relative to
the defrost and opening cycles; redefine the measurand as
``temperature at the middle shelf centre, sampled in the steady
state between defrost cycles, with the door unopened for
$\geq 1$ hour'' to make the target achievable.

\begin{exercise}
Design a comparison protocol for a clinical-laboratory measurement
of blood glucose by two methods. Given an acceptance criterion
supplied by the problem statement (for example, $\pm 15\%$ of the
reference value at glucose concentrations above
$5.55 \,\si{\milli\mol\per\liter}$), state how the laboratory would
establish whether the two methods agree within their combined
uncertainty, and what investigation follows when they disagree.
\end{exercise}

\paragraph{Discussion.}
Protocol: measure a representative panel of $\geq 40$ patient
specimens spanning the clinical range ($2$ to $25 \,\si{\milli
\mol\per\liter}$) by both methods within a short time window
(same blood draw, same day). For each specimen compute the
difference $\Delta = m_{1} - m_{2}$ and its combined standard
uncertainty $u(\Delta) = \sqrt{u_{1}^{2} + u_{2}^{2}}$. The
methods agree at the specimen level if
$|\Delta| \leq 2 u(\Delta)$; at the cohort level, plot Bland-
Altman differences and confirm the mean bias is within the
acceptance criterion across the clinical range. When two methods
disagree, the investigation lane is: (i) re-measure the disagreeing
specimens to rule out single-sample artefacts; (ii) check
calibration of both methods against a certified reference material;
(iii) test for matrix effects (interferents such as haemolysis,
lipaemia); (iv) examine the time difference between the two
methods' measurements for analyte degradation. Cite the GUM for
the combined-uncertainty framework
\cite{std:iso-iec-guide98}.

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A laboratory reports a measurement of $5.32 \pm 0.001 \,\si
{\volt}$. The instrument is a digital multimeter rated at $\pm
0.05\,\%$ of reading. State whether the reported uncertainty is
plausible, and identify the most likely error in the report.
\end{exercise}

\paragraph{Solution.}
The exercise is a diagnostic question: given a reported value and
uncertainty, identify whether the report's number is plausible
against the instrument's documented characteristics, and locate
the report's most likely error.

\emph{Step 1: extract the manufacturer's accuracy specification.}
A digital multimeter rated $\pm 0.05\%$ of reading at
$5.32\,\si{\volt}$ has a specification-derived contribution of
$0.0005 \times 5.32 \approx 2.7\,\si{\milli\volt}$. The
specification is conventionally a $k = 2$ expanded uncertainty for
laboratory bench DMMs (the manufacturer's interpretation should
be confirmed against the data sheet's coverage convention).

\emph{Step 2: convert to standard uncertainty.} At $k = 2$, the
standard uncertainty from the specification alone is approximately
$1.3\,\si{\milli\volt}$. This is a lower bound: the actual
standard uncertainty for an instrument in service includes
additional contributions (last-digit resolution, repeatability,
environmental variation, sampling and bias) that compose with the
specification term in quadrature.

\emph{Step 3: compare to the reported value.} The reported
$\pm 0.001\,\si{\volt} = 1\,\si{\milli\volt}$ is smaller than the
specification-alone standard uncertainty of $1.3\,\si{\milli\volt}$.
This is the diagnostic signal: the report claims tighter
uncertainty than the instrument's documented accuracy allows. The
report is implausible.

\emph{Step 4: enumerate likely error mechanisms.}
\begin{itemize}[leftmargin=*]
  \item \emph{Resolution-only interpretation}: the reporter took
    the last-digit resolution (typically $1\,\si{\milli\volt}$ for
    a 5-digit DMM displaying $X.XXX\,\si{\volt}$) as the
    uncertainty, ignoring the accuracy-spec contribution. This is
    a common rookie error and matches the reported value exactly.
  \item \emph{Standard-deviation-of-mean only}: the reporter
    averaged many readings, took the standard deviation of the
    mean as the uncertainty, and omitted the systematic accuracy
    contribution. Averaging reduces the random component but
    cannot reduce the calibration offset; the report should include
    both.
  \item \emph{Wrong coverage factor}: the reporter quoted
    $u/(k=2)$ as the standard uncertainty without recognising
    that the manufacturer's specification was already at $k = 2$,
    effectively dividing by $k$ twice.
  \item \emph{Uncertainty mis-quoted as resolution}: the reporter
    confused the display's last-digit position with a statement of
    uncertainty.
\end{itemize}

\emph{Step 5: the working engineering response.} Request the
uncertainty budget: the report's standard uncertainty should be
traceable to its Type A and Type B contributions, with the
manufacturer-spec component visible. The budget will expose which
of the four mechanisms above is at play. The diagnostic question is
not "is the reported number wrong" but "where is the missing
contribution".

\emph{Step 6: a follow-up check.} Even if the reporter corrects
the error, the working reader should verify the reported standard
uncertainty against the application's tolerance and against the
instrument's stated repeatability. A reported uncertainty that is
plausible against the data sheet may still be insufficient for the
downstream application; the audit is not just "is the report
correct" but "is the report fit for purpose".

\begin{exercise}
A pressure measurement gives $P = 100 \,\si{\kilo\pascal}$ with
$u(P) = 0.5 \,\si{\kilo\pascal}$ at $k = 1$. The same pressure
measured by an independent method gives $P = 102 \,\si{\kilo\pascal}$
with $u(P) = 0.4 \,\si{\kilo\pascal}$ at $k = 1$. State whether
the two measurements are consistent within their combined
uncertainty. If not, propose three plausible mechanisms for the
disagreement.
\end{exercise}

\paragraph{Solution.}
The consistency test for two independent measurements of the same
quantity is the normalised-difference criterion. The solution
applies it and then enumerates the typical failure modes when the
test rejects.

\emph{Step 1: compute the difference and its uncertainty.} The
two readings differ by $\Delta = P_{2} - P_{1} = 102 - 100 =
2\,\si{\kilo\pascal}$. The uncertainty of the difference,
assuming the two measurements are independent and at $k = 1$, is
\[
u(\Delta) = \sqrt{u(P_{1})^{2} + u(P_{2})^{2}} =
\sqrt{0.5^{2} + 0.4^{2}} \approx 0.64\,\si{\kilo\pascal}.
\]

\emph{Step 2: compute the normalised difference.} The ratio of
$|\Delta|$ to $u(\Delta)$ is the standardised statistic for
testing the null hypothesis that the two measurements come from
the same population value:
\[
z = \frac{|\Delta|}{u(\Delta)} = \frac{2}{0.64} \approx 3.1.
\]
At $k = 2$ (the typical engineering significance level
corresponding to $\approx 95\%$ coverage for Gaussian errors), the
threshold is $|z| \leq 2$. At $z = 3.1$, the two readings are
inconsistent at the $k = 2$ level; the disagreement is unlikely
to be chance variation alone.

\emph{Step 3: state the conclusion.} The two measurements
\emph{are not consistent} within their combined uncertainty. The
disagreement is large enough to require investigation; treating
the average of the two values as "the answer" would inherit the
unidentified error mechanism.

\emph{Step 4: enumerate plausible disagreement mechanisms.}
\begin{itemize}[leftmargin=*]
  \item \emph{Calibration offset in one method}: one of the two
    instruments has a systematic bias not captured in its stated
    uncertainty. The bias may have appeared since the last
    calibration (drift, mechanical shock, environmental
    excursion) or may have been present at calibration but not
    documented.
  \item \emph{Different quantities measured}: the two methods
    measure subtly different physical quantities that the reporter
    treated as identical. Common cases: absolute vs gauge pressure
    (the difference is one atmosphere if one method references
    vacuum and the other references local atmospheric); fluid
    pressure at the gauge vs at a reference depth (a hydrostatic
    column adds pressure); temperature-corrected vs uncorrected
    reading (pressure transducers are temperature-dependent).
  \item \emph{Method-specific bias}: each method has its own
    operating envelope and may be measuring outside it for one of
    the two readings. A diaphragm pressure transducer near its
    burst rating shows nonlinear response; a deadweight tester
    requires a level reference plane.
  \item \emph{Stated uncertainties incomplete}: one or both
    reports omit a Type B contribution (operator effect,
    environmental variation, calibration interval not respected).
    The actual combined standard uncertainty exceeds the reported
    $u(P_{1})^{2} + u(P_{2})^{2}$.
  \item \emph{Sample heterogeneity}: in some applications the
    "same pressure" assumption is itself faulty. Two transducers
    measuring "the same" pressure at different tap locations or
    different times may see real differences from spatial or
    temporal variation in the system.
\end{itemize}

\emph{Step 5: the working engineering response.} The chapter's
discipline applies directly: widen the reported uncertainty band
to encompass both readings until the disagreement mechanism is
identified. A combined report might state $P = 101 \pm 2\,\si{
\kilo\pascal}$ with the note "two independent methods disagree;
investigation pending". Refusing to issue any number is sometimes
the right choice; an unjustified average is rarely.

\emph{Step 6: the lesson.} The exercise reproduces in miniature the
chapter's failure section: two independent measurements that
disagree by more than their combined uncertainty are evidence the
uncertainty budget is incomplete. The Hubble case is the
high-stakes version of the same diagnostic. The working
investigator's first move is the variance-of-the-difference test,
followed by the enumeration of mechanisms above. The corrective
action depends on which mechanism applies: a cross-check of both
instruments against a reference standard at known pressure
isolates the calibration-offset case from the different-quantity
case.

\begin{exercise}
A spreadsheet computes $y = (x_{1} + x_{2}) / x_{3}$ from inputs
with stated uncertainties $u_{1}, u_{2}, u_{3}$. The spreadsheet
reports the standard uncertainty of $y$ as $u_{1} + u_{2} + u_{3}$
divided by $x_{3}$. Identify the error and write the correct
propagation formula.
\end{exercise}

\paragraph{Solution.}
The spreadsheet's formula contains two structurally different
errors. The diagnosis identifies each and gives the correct form.

\emph{Step 1: identify the linear-sum error.} The spreadsheet
writes $u_{1} + u_{2} + u_{3}$, a linear sum of the three input
uncertainties. The correct combination for independent inputs is
the root-sum-square: $\sqrt{u_{1}^{2} + u_{2}^{2} + u_{3}^{2}}$
for additive terms; the linear sum always overstates the combined
uncertainty by up to a factor of $\sqrt{n}$ for $n$ equal
contributions ($\sqrt{3}$ at $n = 3$ for three equal terms).

\emph{Step 2: identify the division-rule error.} The spreadsheet
divides by $x_{3}$ at the end, treating $u_{3}$ as an additive
contribution that is then scaled. The correct propagation through
division uses the product-of-powers rule: for $y = s/x_{3}$, the
\emph{relative} uncertainty satisfies $(u(y)/y)^{2} =
(u(s)/s)^{2} + (u_{3}/x_{3})^{2}$. The division enters as a
relative-uncertainty contribution, not as a scaling of the
absolute uncertainty.

\emph{Step 3: write the correct formula.} Let $s = x_{1} + x_{2}$
with combined absolute uncertainty $u^{2}(s) = u_{1}^{2} + u_{2}^{2}$.
Then $y = s/x_{3}$ has relative uncertainty
\[
\left(\frac{u(y)}{y}\right)^{2} =
\left(\frac{u(s)}{s}\right)^{2} +
\left(\frac{u_{3}}{x_{3}}\right)^{2},
\]
and absolute uncertainty
\[
u(y) = |y| \sqrt{(u(s)/s)^{2} + (u_{3}/x_{3})^{2}}.
\]
This is the GUM-compliant working result.

\emph{Step 4: quantify the spreadsheet's error.} For three equal
relative uncertainties of $1\%$ each, the spreadsheet's formula
gives an absolute uncertainty of approximately $(0.03)(s + u_{3}
x_{3}^{-1})/x_{3}$ or some variant, depending on the literal
interpretation; the correct value is $|y|\sqrt{3 \times 0.01^{2}}
\approx 1.73\% \cdot |y|$ relative. The spreadsheet's
linear-sum part overstates by up to $\sqrt{3}$; its division
treatment is structurally wrong and the magnitude of its error
depends on $s$ and $x_{3}$.

\emph{Step 5: identify the broader category of spreadsheet
errors.} Linear sums in place of root-sum-square, division-by-mean
in place of relative-uncertainty division, and ignored covariance
between cells that share a common input are the three most common
spreadsheet propagation errors. A spreadsheet that propagates
uncertainty in production should be audited against the GUM rules
either by inspection or by Monte Carlo simulation against the
spreadsheet's actual cell-arithmetic.

\emph{Step 6: the corrective workflow.} Replace the offending
formula with the correct expression. Add a row to the uncertainty
budget for each intermediate quantity ($s, y$) and verify each
contribution against the GUM rules. For production use,
documenting the cell formulas alongside the budget supports the
audit, and a Monte Carlo cross-check at $10^{4}$ to $10^{5}$
samples confirms the closed-form result.

\begin{exercise}
A torque-wrench specification claims $\pm 4\%$ at $k = 1$. The
manufacturer tests the wrench against a reference and the
measurements scatter at $\pm 4\%$ at $k = 2$. State whether the
wrench meets its specification and what additional information
the user needs to make the comparison.
\end{exercise}

\paragraph{Solution sketch.}
The spec is $4\%$ at $k = 1$, so the standard uncertainty must be
$\leq 4\%$. The test reports $\pm 4\%$ at $k = 2$, so the
measured standard uncertainty is $2\%$. The wrench passes the
specification with room to spare. Additional information needed:
(i) the distribution underlying the test data (the $k$ conversion
assumes a Gaussian shape); (ii) whether the test included any
systematic offset between wrench and reference, or only random
scatter; (iii) whether the $\pm 4\%$ in the spec includes both
calibration drift over the recalibration interval and
operator-dependent reading error, or only the wrench's intrinsic
accuracy.

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
The Hubble Space Telescope's primary-mirror figure measurement
relied on the reflective null corrector while two auxiliary
optical tests indicated discrepancies. Argue, in 200 to 300
words, what the reported uncertainty on the mirror's figure should
have been if the auxiliary-test results had been incorporated.
Cite the OSBI report \cite{acc:nasa-hubble-optical-systems-1990}.
\end{exercise}

\paragraph{Discussion.}
A rubric expects the argument to address: (i) the difference
between the reflective null corrector reading and the auxiliary
inverse-null and refractive-null readings was sizable; the
discrepancy was a measurement signal, not noise. (ii) When
independent measurement methods disagree by more than their
individually-reported uncertainties, the engineering response is
to widen the reported uncertainty to encompass the disagreement,
or to identify which method is wrong by an independent third
measurement, or to halt fabrication until the cause of the
disagreement is identified. (iii) The OSBI traced the
$1.3 \,\si{\milli\meter}$ misposition through to a $2 \,\si
{\micro\meter}$ edge figure error; with auxiliary-test
incorporation the reported figure uncertainty would have been
multiple wavelengths of light, not the diffraction-limited figure
the project assumed. (iv) The institutional failure was treating
the reflective null corrector as authoritative because of cost
and schedule pressure, not because of an evidence-based
preference. Cite the OSBI report directly. Strong responses
distinguish between the technical question (what should the
uncertainty have been?) and the organisational question (why was
the smaller uncertainty reported anyway?).

\begin{exercise}
A polling firm reports a $\pm 3\%$ margin of error on a poll of
$1{,}000$ respondents and concludes that two candidates are
"statistically tied" within the margin. Argue, in 200 to 300
words, whether the stated margin of error correctly characterises
the poll's uncertainty, and identify at least three systematic
sources that the margin does not include.
\end{exercise}

\paragraph{Discussion.}
The $\pm 3\%$ margin of error is the sampling-only component:
$1/\sqrt{1000} \approx 3.2\%$, which is the standard error for
a proportion near $0.5$ at $k = 2$. Systematic sources the margin
omits include: (i) non-response bias (those who answer differ
systematically from those who do not); (ii) coverage bias (whom
the sampling frame can reach: landline, mobile, web, mailing
list, social media); (iii) question wording and order effects;
(iv) weighting and post-stratification choices; (v) likely-voter
modelling assumptions for election polls; (vi) social-desirability
bias on sensitive items; (vii) sample frame currency. Each of
these can shift the reported proportion by several percentage
points. Empirical studies of polling error in elections show
total error commonly exceeds the sampling margin by a factor of
two to three. The ``statistically tied'' conclusion is correct
only under the assumption that the sampling component is the
total uncertainty; under realistic assumptions about systematic
sources, the report should widen the comparison band before
declaring a tie. The framework parallels the GUM's Type A
versus Type B distinction: the margin of error is the Type A
component computed from the sample, and the omitted systematic
sources are the Type B components that the report did not
evaluate.

\begin{exercise}
A regulatory agency reports a cost-benefit analysis with a single
point estimate of net benefit of $\$1.2$ billion. The underlying
analysis produced a range from $\$0.4$ to $\$2.1$ billion. State
what the single point estimate communicates that the range does
not, and identify the discipline that would prevent the
single-estimate report.
\end{exercise}

\paragraph{Discussion.}
The single point estimate communicates false precision: a
decision-maker reading $\$1.2$ billion may treat the number as
known to a few percent, when the underlying analysis admits
factor-of-two variation. The range $\$0.4$ to $\$2.1$ billion
implies that under unfavourable but plausible assumptions the
net benefit could be only a third of the headline; the regulator
needs that information to decide whether to defer, modify, or
phase the regulation. The discipline that prevents the
single-estimate report is the requirement to report central
estimate \emph{and} interval together: $\$1.2 \,[\$0.4, \$2.1]$
billion at a stated confidence level, with a sensitivity table
showing which input assumptions drive the lower and upper bounds.
US OMB Circular A-4 (2003, updated 2023) requires precisely this
structure for major federal rulemakings; many agency reports
still collapse the range to a point for press release. A reviewer
in this lane should always ask: what does the interval look
like, and where is the sensitivity table?

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A senior colleague reports a measurement as $5.32 \,\si{\volt}$
with no uncertainty. Write a one-paragraph response that takes the
colleague seriously, identifies the specific information missing
to interpret the result, and proposes a minimum acceptable form
for the report.
\end{exercise}

\paragraph{Discussion.}
A rubric expects: respectful tone (the colleague is presumed to
have measured carefully, not been careless); concrete request for
the standard or expanded uncertainty, the coverage factor, the
instrument and its calibration date, the number of replicate
readings or the single-reading basis, and the environmental
conditions; a proposed minimum form such as $5.32 \pm 0.05 \,\si
{\volt}$ at $k = 2$ with stated coverage; a closing offer to
help compute the uncertainty budget if the colleague has the
underlying data but has not yet derived the figure. Avoid: blame,
gatekeeping, or framing the request as criticism of the
colleague's competence. The discipline here is the
working-engineer practice of not using under-specified numbers in
decisions.

\begin{exercise}
A startup is preparing a regulatory submission and the regulator
expects a $\pm 5\%$ uncertainty on a key measurement. The startup's
data supports $\pm 8\%$. Argue, in 200 to 300 words, what the
startup should do, why narrowing the reported uncertainty to
$\pm 5\%$ would be a discipline failure even if it satisfies the
regulator, and what the startup's options are.
\end{exercise}

\paragraph{Discussion.}
The startup should report what the data supports. Narrowing the
uncertainty to satisfy the regulator is a discipline failure for
three reasons: (i) the reported figure becomes a misrepresentation
of the measurement evidence, the same failure mode that the
Volkswagen defeat-device case carried at regulatory scale; (ii)
post-submission re-measurement or independent verification will
expose the discrepancy, with consequences ranging from rejected
submission to fraud findings; (iii) the regulatory $\pm 5\%$
expectation is usually a derived requirement based on the
intended use; if the data cannot meet it, the underlying decision
must be reconsidered, not the reported number. The startup's
options are: (a) collect additional data through more replicate
measurements, a more accurate instrument, or a better-controlled
procedure to reach $\pm 5\%$ honestly (cost and schedule
implications); (b) submit at $\pm 8\%$ with a transparent
uncertainty budget and a request to defer the submission or to
relax the requirement; (c) re-scope the application so the $\pm
8\%$ figure is sufficient for the regulator's purpose. The
discipline-preserving paths are (a) and (c), with (b) preserving
honesty even if the regulator's response is unfavourable.

\begin{exercise}
Argue, in one paragraph, whether engineering training should
emphasise random or systematic error reduction. Take a position
and defend it with a specific example from your own experience or
a documented case.
\end{exercise}

\paragraph{Discussion.}
Either position is defensible if argued with a concrete case.
A defensible answer: emphasise systematic-error training, because
random-error reduction is procedurally simple (average more) and
arithmetically capped by the same $1/\sqrt{n}$ law every reader
will encounter, while systematic-error identification is harder,
requires understanding the instrument and procedure, and is the
mechanism behind most public engineering failures (Mars Climate
Orbiter unit error, Hubble null-corrector misposition, Therac-25
race condition, Boeing MCAS sensor logic). The opposite position
is also defensible: random-error training equips the working
engineer to read laboratory data without confusion, since
systematic errors are rarer in routine work and need a teacher's
help to identify. A weak answer chooses ``both equally'' without
distinguishing teaching emphases. Strong answers cite a specific
case from the named-cases registry or from the reader's
experience.

\begin{exercise}
A data sheet states `accuracy: $\pm 0.8\%$ of reading' with no
coverage factor and no distribution. Write the uncertainty-budget
entry three ways: as a rectangular half-width, as a $k = 2$ Gaussian
expanded uncertainty, and as an unresolved specification requiring
manufacturer clarification. Compare the resulting standard
uncertainties and state which assumption a conservative engineer
would use absent further information.
\end{exercise}

\paragraph{Discussion.}
Reading $R$, accuracy half-width $a = 0.008 R$. (i) Rectangular:
$u = a/\sqrt{3} \approx 0.00462 R$. (ii) Gaussian at $k = 2$:
$u = a/2 = 0.004 R$. (iii) Unresolved: record the spec verbatim
and add a note that manufacturer clarification is required
before propagation. The two computed standard uncertainties
differ by about $15\%$. A conservative engineer uses the
rectangular assumption absent further information: it is the GUM
default for a single bounded specification with no distribution
stated \cite{std:iso-iec-guide98} and is the larger of the two
plausible interpretations. The most defensible practice combines
both: report the rectangular interpretation as the budget entry,
attach the data sheet excerpt, and flag the resolution as an
open item.

\begin{exercise}
Run a Monte Carlo propagation for $y = x_{1} - x_{2}$ with equal
standard uncertainties and correlations $\rho = 0$, $0.5$, $0.9$,
and $0.99$. Compare the empirical standard deviation with the
correlated-difference formula
$u_{c}^{2}(y) = u^{2}(x_{1}) + u^{2}(x_{2}) - 2\rho u(x_{1})
u(x_{2})$, and explain why common-mode cancellation depends on
positive correlation between the two readings.
\end{exercise}

\paragraph{Solution sketch.}
With $u(x_{1}) = u(x_{2}) = 1$, the formula gives:
$\rho = 0$: $u(y) = \sqrt{2} \approx 1.41$;
$\rho = 0.5$: $u(y) = 1$;
$\rho = 0.9$: $u(y) \approx 0.45$;
$\rho = 0.99$: $u(y) \approx 0.14$. Monte Carlo with $10^{5}$
samples reproduces these within $1\%$. The mechanism: positive
correlation means $x_{1}$ and $x_{2}$ fluctuate together (when
$x_{1}$ is above its mean, $x_{2}$ tends to be above its mean
too). The shared fluctuation cancels in the difference; only the
uncorrelated residual contributes to $u(y)$. Negative correlation
would amplify the difference instead. Reference implementation in
\texttt{code/common\_mode.py}.
