\chapter{Genetics, transcription, translation}
\label{vol06:ch03}

\epigraph{It has not escaped our notice that the specific pairing we
have postulated immediately suggests a possible copying mechanism for
the genetic material.}{James D. Watson and Francis H. C. Crick,
\emph{Nature} 171, 737 (1953), closing paragraph
\cite[p.~738]{paper:watson-crick1953}.}

\chapmeta{Half-life: foundational. Archetypes: information, failure, scaling. Exercise target: 30. Project track: simulation of genetic drift in a finite population (Wright-Fisher model). Hazard class: none (in-silico work). Reader path default: standard, with sections explicitly tagged. Named cases: none required; cancer and hereditary-disease examples are described as mechanism classes.}

\noindent
A cell that does not replicate dies in one generation. A cell that
replicates without controlled error dies in many generations through
the accumulation of damage. The chapter develops the molecular
machinery by which cells store, copy, transcribe, and translate
genetic information, treats the genetic code as an engineering
specification, quantifies the error rates of replication and repair,
surveys the sequencing technologies that make modern molecular
engineering possible, and closes on the failure modes that arise when
the information machinery breaks (cancer, hereditary disease, ageing).

\section{DNA as digital storage}\pathtag{core}

A cell's genetic information is stored in deoxyribonucleic acid (DNA),
a double-helical polymer of four nucleotides. Each nucleotide consists
of a phosphate group, a deoxyribose sugar, and one of four nitrogenous
bases: adenine (A), thymine (T), guanine (G), or cytosine (C). The
double helix consists of two antiparallel strands held together by
hydrogen bonds between complementary base pairs: A pairs with T
(two hydrogen bonds), G pairs with C (three hydrogen bonds). The
geometry of the helix is set by the sugar-phosphate backbone and the
stacking of the base pairs; the canonical B-form has a diameter of
$2\,\text{nm}$, a rise of $0.34\,\text{nm}$ per base pair, and a
helical pitch of $3.4\,\text{nm}$ (ten base pairs) per turn
\cite{paper:watson-crick1953}.

The information content per base pair is $\log_2 4 = 2\,\text{bits}$.
The human genome is $3.2 \times 10^9\,\text{base pairs}$ in each haploid
set, or $6.4 \times 10^9\,\text{bits} = 0.8\,\text{GB}$ per diploid
nucleus. Compressed by removing repetitive sequence, the irreducible
information content is closer to $400$ to $700\,\text{MB}$. The
\emph{E.~coli} genome is $4.6 \times 10^6$ base pairs, or
$1.15\,\text{MB}$. The bacteriophage $\lambda$ has $48\,\text{kbp}$
($12\,\text{kB}$); a small viroid (a non-coding plant RNA) has
$250$ to $400\,\text{bases}$ ($60$ to $100\,\text{bytes}$). The
information densities cover seven orders of magnitude across the
biosphere.

\engterm{The molecular engineering of the helix}. The double helix is
mechanically stiff at lengths below its persistence length
$L_p \approx 50\,\text{nm}$ ($\approx 150\,\text{bp}$) and flexible
at longer lengths. The persistence length sets the geometric
limitations on how the genome can fold and how it can be threaded
through proteins. A $3.2 \times 10^9\,\text{bp}$ human genome
stretched end to end would be $1\,\text{m}$ long; the cell packages it
into a $5\,\si{\micro\meter}$ nucleus by wrapping the DNA around
histone proteins to form nucleosomes (each nucleosome is a
$\sim 11\,\text{nm}$ disc wrapping $147\,\text{bp}$), assembling
nucleosomes into a $30\,\text{nm}$ fibre, and folding the fibre into
higher-order loops and domains. The packaging factor is
$\sim 10^4$ in interphase chromatin and approximately $10^5$ in
metaphase chromosomes. The packaging is dynamic: the cell unwinds
regions of chromatin to allow access to transcription and
replication machinery, and re-winds them when access is no longer
needed. Chromatin structure is a control variable in gene expression
that Section~3.3 will return to.

\engterm{Why DNA, not RNA}. The cell uses DNA for stable information
storage and RNA for working transcripts. The choice is not arbitrary:
the absence of the $2'$-hydroxyl in deoxyribose makes DNA
substantially more chemically stable than RNA (the $2'$-OH catalyses
backbone hydrolysis in RNA), and the use of thymine rather than
uracil allows the cell to distinguish original-sequence thymines from
cytosine deamination products (cytosine deaminates spontaneously to
uracil at a measurable rate; if the genome used uracil, the deaminations
would be invisible to repair machinery). The substitution of thymine
for uracil is one of the cell's most elegant engineering choices: a
single methyl group provides the error-detection signature that
makes large-genome heredity feasible.

\engterm{Replication}. The double helix is replicated by separating
the two strands and using each as a template for synthesis of a new
complementary strand. The reaction is catalysed by DNA polymerases
that add nucleotides to a growing $3'$ end at a rate of approximately
$1000\,\text{nucleotides per second}$ in bacterial replication and
$100$ to $1000$ in eukaryotic replication. Bacterial replication is
bidirectional from a single origin (oriC in \emph{E.~coli}); eukaryotic
replication is bidirectional from thousands of origins distributed
along each chromosome. The leading strand is synthesised continuously;
the lagging strand is synthesised discontinuously in short
\engterm{Okazaki fragments} that are then joined by ligase. The
machinery (helicase, topoisomerase, primase, polymerase, sliding clamp,
ligase, and a dozen other accessory proteins) assembles into a
replication fork that resembles, in its engineering, a coordinated
multi-tool assembly machine.

\engterm{Replication errors and proofreading}. DNA polymerase
incorporates the wrong nucleotide approximately once per $10^5$
incorporations from base-pairing fidelity alone. The polymerase's
$3' \rightarrow 5'$ exonuclease domain removes most of these errors
by proofreading immediately after incorporation, reducing the error
rate to approximately $10^{-7}$ per nucleotide. Mismatch-repair
machinery downstream catches most of the remaining errors, reducing
the error rate at fixation to approximately $10^{-9}$ to $10^{-10}$
per nucleotide per generation for fast-growing bacteria
\cite[ch.~25]{text:lehninger-biochemistry}. The cascade (base
selection, proofreading, mismatch repair) is an example of a
multi-stage error-correction architecture whose effective error rate
is the product of the per-stage error rates. Engineering parallels
in computer storage (parity, ECC, RAID) follow the same architectural
pattern.

\begin{principle}[Genetic information is digital and error-corrected]
The cell stores information in a four-symbol digital code with a
multi-stage error-correction architecture whose effective per-base
error rate ($10^{-9}$ to $10^{-10}$) is several orders of magnitude
smaller than the per-stage error rate of any single step. The
architecture is the engineering condition that makes genomes of
$10^9$ to $10^{11}$ base pairs faithfully heritable across
generations. A reduction of the error-correction reliability by a
single factor of ten (loss of one proofreading or mismatch-repair
stage) produces measurable disease (the \engterm{mutator phenotype}
in cancer) within a few cell generations.
\end{principle}

\section{The central dogma engineering-first}\pathtag{core}

The \engterm{central dogma}, due to Francis Crick in 1957, states
that information in a cell flows in one direction: DNA $\rightarrow$
RNA $\rightarrow$ protein. The dogma has well-known refinements
(reverse transcription from RNA to DNA in retroviruses; replication of
RNA viruses) but the engineering reading remains useful: the cell
stores its information as DNA, transcribes it to RNA when needed, and
translates the RNA into the protein that performs the cellular work.

\engterm{Transcription}. The cell copies a specific stretch of DNA
into a complementary RNA strand by means of RNA polymerase. The
polymerase binds to a \engterm{promoter} sequence upstream of the
gene, unwinds the DNA locally, and synthesises RNA at a rate of
approximately $50$ to $100\,\text{nucleotides per second}$. The RNA
strand is identical in sequence to the non-template DNA strand, with
uracil replacing thymine. Transcription terminates at a
\engterm{terminator} sequence (in bacteria) or by polyadenylation
signal (in eukaryotes). Bacterial and eukaryotic transcription differ
in detail (bacteria use a single RNA polymerase with sigma factors
for promoter recognition; eukaryotes use three distinct polymerases
plus a complex of general transcription factors at each promoter) but
share the engineering structure: a regulated synthesis of a
sequence-defined RNA copy from a sequence-defined DNA template.

\engterm{RNA processing}. In eukaryotes, the primary RNA transcript
undergoes three principal modifications before translation. A
$5'$ cap (a modified guanosine) is added to the front; a poly(A) tail
of $\sim 200$ adenines is added to the back; and \engterm{introns}
(non-coding sequences between coding \engterm{exons}) are removed by
the spliceosome. The processed mRNA is exported from the nucleus to
the cytoplasm for translation. Alternative splicing (different
combinations of exons included or excluded) allows a single gene to
encode multiple protein variants; the human genome's
$\sim 20\,000$ protein-coding genes encode an estimated $\sim 100\,000$
distinct protein variants through alternative splicing.

\engterm{Translation}. The mRNA is decoded by ribosomes into a
protein. The ribosome is a large ribonucleoprotein complex
($\sim 2.5\,\text{MDa}$ in bacteria, $\sim 4\,\text{MDa}$ in
eukaryotes) consisting of two subunits with three transfer-RNA
(tRNA) binding sites. Each tRNA carries an amino acid at one end and
a three-nucleotide anticodon at the other; the anticodon base-pairs
with a complementary codon in the mRNA. The ribosome moves along the
mRNA in steps of three nucleotides, with each step adding one amino
acid to the growing peptide chain. Translation proceeds at
$15$ to $20$ amino acids per second in bacteria, slightly slower in
eukaryotes. The ribosome reads the mRNA from the $5'$ end to the
$3'$ end, starts at an AUG start codon, and stops at one of three
stop codons (UAA, UAG, UGA). The completed peptide is released and
folds into a functional protein (Chapter~\ref{vol06:ch04}).

\engterm{Error rates}. Translation makes errors at a rate of
approximately $10^{-3}$ to $10^{-4}$ per codon, dominated by tRNA
mischarging (the aminoacyl-tRNA synthetase loading the wrong amino
acid onto a tRNA) and by codon-anticodon mispairing at the ribosome.
The rate is much higher than the DNA replication rate because the
errors are transient (one mistranslated protein molecule among many
correctly translated copies) and the cost of correcting them at
synthesis time exceeds the cost of tolerating them in the protein
pool. The cell's error-rate budget at each information-processing
step is engineered to the cost of failure at that step: replication
errors are heritable and so are corrected aggressively; translation
errors are transient and so are tolerated more.

\section{Gene expression and regulation}\pathtag{standard}

A cell's genome contains the parts list of every protein the cell
might ever need. The cell does not synthesise all of them at once;
different cell types, different developmental stages, and different
environmental conditions require different protein complements. The
machinery that selects which genes are expressed and at what level is
the \engterm{gene-expression regulatory network}.

\engterm{Transcriptional regulation}. The dominant level of control is
the rate of transcription initiation at each gene. The control is
exerted by \engterm{transcription factors}, proteins that bind
specific DNA sequences near a gene's promoter and either recruit or
exclude the RNA polymerase. A typical bacterial gene's promoter is
regulated by one to a few transcription factors; a typical
eukaryotic gene's promoter is regulated by a dozen or more
transcription factors plus distal enhancer elements that may be tens
of kilobases away. The combinatorial logic of multiple transcription
factors at a single promoter implements logical operations (AND, OR,
NOT) on the regulatory inputs; the network architecture computes
gene expression as a function of the cell's state.

\engterm{Network architecture}. The cell's transcription network has
been mapped exhaustively for \emph{E.~coli} (approximately $300$
known transcription factors regulating $4288$ genes) and for several
other model organisms. The network is sparse: each transcription
factor regulates a few to a few hundred targets; each gene is
regulated by a few transcription factors. The network contains
recurring motifs whose engineering function has been worked out:
auto-regulation (a transcription factor regulating its own gene,
either negatively for noise reduction or positively for switch-like
behaviour), the feed-forward loop (Section~\ref{vol06:ch01} introduced
the type-1 coherent loop, with its delay and noise-filtering behaviour),
the single-input module (a master regulator controlling a set of
co-regulated targets), and the dense overlapping regulon (multiple
factors jointly controlling overlapping target sets)
\cite[ch.~6]{text:alon-systems-biology}.

\engterm{Post-transcriptional regulation}. The cell controls protein
abundance not only by transcription rate but also by mRNA stability
(half-lives ranging from minutes for short-lived mRNAs to days for
stable structural mRNAs), translation efficiency (set by the mRNA's
$5'$ untranslated region and by sequence-specific binding factors),
protein stability (set by the protein's degradation signals,
including ubiquitin tagging for proteasomal degradation), and
post-translational modification (phosphorylation, acetylation,
methylation). The cell's effective gene-expression control surface
combines all these levels, with the result that the steady-state
abundance of a protein can vary by factors of $10^4$ to $10^6$ across
cell states without changing the underlying DNA sequence.

\engterm{Epigenetic regulation}. Some changes in gene expression
persist across cell divisions without changes in DNA sequence. The
two principal mechanisms are DNA methylation (covalent attachment of
methyl groups to cytosines, suppressing gene expression in mammals)
and histone modification (acetylation, methylation, and other
covalent marks on the histone proteins around which DNA is wrapped,
affecting chromatin accessibility). Together these constitute the
\engterm{epigenome}, a layer of regulatory information that is
copied along with the DNA during replication and that maintains
cell-type-specific gene expression patterns. Epigenetic engineering
is an active area of biomanufacturing and therapy at the time of
writing (current as of 2025).

\section{The genetic code}\pathtag{standard}

The \engterm{genetic code} is the mapping from nucleotide triplets
(codons) to amino acids and stop signals. There are $4^3 = 64$
possible codons. Three are stop codons (UAA, UAG, UGA); the remaining
$61$ are sense codons that specify the $20$ standard amino acids.
The code is therefore degenerate: most amino acids are specified by
two to six different codons, and only methionine (AUG, the standard
start codon) and tryptophan (UGG) are specified by a single codon.

\engterm{Code structure}. The degeneracy follows a regular pattern.
The third codon position (the ``wobble position'') is the most
variable; for many amino acids, all four nucleotides at the third
position specify the same amino acid (the third-position degeneracy
is fourfold). The first and second positions are more constrained.
The codon table groups codons by physical and chemical properties of
the encoded amino acid: codons starting with U specify hydrophobic
amino acids (phenylalanine, leucine, isoleucine, methionine, valine);
codons with C in the second position specify small or polar amino
acids; codons with A in the second position specify polar or charged
amino acids. The pattern is not random: a single-nucleotide
substitution is statistically more likely to substitute a chemically
similar amino acid (a conservative substitution) than to substitute a
chemically dissimilar amino acid (a radical substitution). The bias is
the code's principal error-mitigation feature.

\engterm{Universality}. The code is almost universal: the same codon
specifies the same amino acid in essentially every organism on Earth.
The few known exceptions (variant codes in mitochondria, in some
ciliates, in some bacterial lineages) involve a small number of
codons. The universality reflects the evolutionary cost of changing
the code: once a code is in place, any change requires simultaneous
changes in the tRNA pool, the aminoacyl-tRNA synthetases, and every
gene whose codons were affected. The code is therefore an example of
\engterm{frozen accident}: a feature that was variable at the origin
of life but has been frozen since.

\engterm{Designed by evolution}. The genetic code's structure (the
specific assignment of codons to amino acids) is the result of
evolutionary optimisation against several competing pressures:
translation accuracy (similar codons should specify similar amino
acids, so that translation errors are tolerable), mutational
robustness (single-nucleotide mutations should preferentially produce
chemically similar substitutions), and biosynthetic cost (frequently
used amino acids should have codons that minimise the cell's resource
demand on tRNA synthesis). The standard genetic code performs better
on these measures than the vast majority of random codes
\cite[ch.~27]{text:lehninger-biochemistry}. The reading is that the
code was selected by evolution as an engineering trade-off, not
designed by deliberate optimisation.

\section{Mutation, repair, and error rates}\pathtag{core}

\engterm{Mutations}. A mutation is a change in the DNA sequence. The
classes:

\begin{itemize}[leftmargin=*]
\item \engterm{Point mutations} (single base changes), which divide
  further into transitions (purine $\leftrightarrow$ purine, or
  pyrimidine $\leftrightarrow$ pyrimidine) and transversions (purine
  $\leftrightarrow$ pyrimidine). Most spontaneous point mutations are
  transitions because the chemical mechanisms (cytosine deamination,
  guanine oxidation) preferentially produce transitions.
\item \engterm{Insertions and deletions} (indels) of one or more
  base pairs. Indels in coding regions cause frameshift mutations
  unless the indel length is a multiple of three; frameshifts
  typically destroy gene function.
\item \engterm{Large structural variants}: duplications, inversions,
  translocations, copy-number variants. These arise during recombination
  or by repair of double-strand breaks. They are individually rare but
  important contributors to genetic disease and to cancer.
\end{itemize}

\engterm{Mutation rates}. The spontaneous mutation rate per base pair
per generation is approximately $10^{-9}$ in bacteria, $10^{-10}$ in
unicellular eukaryotes, and $10^{-8}$ to $10^{-9}$ in mammalian
germline cells \cite[ch.~25]{text:lehninger-biochemistry}. Per genome
per generation, this gives roughly $0.005$ mutations per bacterial
generation, $0.003$ for yeast, and $\sim 100$ for the human germline.
The human-genome value translates to approximately $50$ to $60$ new
single-base variants per child compared to the parents, a number
that genomic sequencing now confirms directly.

\engterm{DNA damage and repair}. Beyond replication errors, the cell's
DNA is continuously damaged by chemical and physical agents: oxidation
of bases (the most common spontaneous lesion, with approximately
$10\,000$ oxidative lesions per human cell per day), deamination,
alkylation, and ultraviolet-induced pyrimidine dimers. The cell repairs
these lesions through several pathways: base excision repair (for
oxidised and small chemical lesions), nucleotide excision repair (for
bulky lesions like UV-induced dimers), mismatch repair (for replication
errors that escape proofreading), and double-strand-break repair by
homologous recombination or non-homologous end joining (for the
most severe lesions). Each pathway has its own engineering
characteristics (speed, fidelity, applicability), and the cell selects
the appropriate pathway based on the lesion type and the cell-cycle
phase.

\engterm{Error rates in engineering context}. The cell's
$10^{-9}$ to $10^{-10}$ per-base error rate is comparable to the
bit-error rate of high-quality hard disk storage
($\sim 10^{-15}$ with ECC, but $\sim 10^{-6}$ without). The cell's
multi-stage architecture (replication accuracy, proofreading,
mismatch repair, damage repair) achieves its low overall error rate
by composition of imperfect stages, the same architecture pattern as
modern storage systems. The engineering reading is that DNA storage,
treated as an engineered system, is one of the most reliable
information-storage technologies ever produced, with retention times
of $10^4$ to $10^6$ years under appropriate conditions (the recovery
of ancient DNA from specimens up to $10^6$ years old demonstrates the
shelf life).

\section{Modern sequencing technology}\pathtag{standard}

The ability to read DNA sequences has transformed biology and
biotechnology over the past two decades. The chapter surveys the
principal technologies in the order they have emerged.

\engterm{Sanger sequencing}. The \engterm{chain-termination method}
developed by Frederick Sanger in 1977 used DNA polymerase to
synthesise a complementary strand in the presence of low concentrations
of fluorescently labelled dideoxynucleotides. Each ddNTP terminates
the chain upon incorporation; the resulting mixture of variable-length
fluorescently labelled fragments is separated by electrophoresis and
the sequence read from the fragment lengths. Sanger sequencing achieves
read lengths of $500$ to $1000$ bases at error rates of $\sim 10^{-4}$,
and remained the dominant technology from $1977$ to approximately
$2007$. The Human Genome Project (1990 to 2003) was completed with
Sanger sequencing at a final cost of approximately $\$3$ billion for
the first reference genome.

\engterm{Short-read next-generation sequencing}. From 2005 onward,
Illumina's reversible-terminator chemistry (and earlier Solexa
technology) drove sequencing costs down by approximately six orders
of magnitude. The platform sequences millions of short ($50$ to
$300\,\text{base}$) reads in parallel by cycling fluorescent
nucleotides through a flow cell of immobilised DNA fragments. Error
rates are approximately $10^{-3}$, mostly substitutions; throughputs
are $10^{12}$ to $10^{13}$ bases per run. The cost per human genome
fell from $\$3\,\text{billion}$ in 2003 to $\$10\,000$ in 2010 to
$\$1000$ in 2014 to under $\$200$ in 2023 (current as of 2025). The
cost trajectory has substantially outpaced Moore's law in the same
period \cite[ch.~12]{text:alon-systems-biology}.

\engterm{Long-read sequencing}. Two technologies (Pacific Biosciences
single-molecule real-time sequencing and Oxford Nanopore Technologies)
produce reads of $10\,\text{kb}$ to $100\,\text{kb}$ at error rates of
$10^{-2}$ to $10^{-3}$. The long reads resolve repetitive sequences,
structural variants, and full-length transcripts that short-read
sequencing cannot. The combination of long-read scaffolding and
short-read polishing now produces near-complete telomere-to-telomere
human-genome assemblies (the T2T-CHM13 reference, 2022, was the first
complete human genome assembled with this approach).

\engterm{Engineering applications}. Modern sequencing supports
biomanufacturing (quality control of engineered strains), medicine
(diagnostic genome sequencing, oncology genotyping, infectious-disease
identification), and basic biology (genome-wide association studies,
single-cell transcriptomics, microbiome surveys). Chapter~\ref{vol06:ch11}
on bioinformatics develops the algorithmic side of sequence analysis;
Chapter~\ref{vol06:ch09} develops the synthetic-biology applications.

\section{Failure: replication errors, cancer, hereditary disease}\pathtag{core}

The cell's information-processing machinery fails in characteristic
ways. The failures are mechanism-class, not single named events. We
identify the principal classes.

\engterm{Cancer as a failure of replication control}. Cancer is the
disease of unregulated cell proliferation. Its molecular basis is the
accumulation of mutations in genes that control proliferation, cell
death, and genomic integrity. Two functional classes of cancer gene
are recognised. \engterm{Oncogenes} are normal cell-growth-promoting
genes (proto-oncogenes) whose activated, mutated forms drive
proliferation; activating mutations include point mutations
(\emph{KRAS}, \emph{BRAF}), gene amplifications (\emph{MYC},
\emph{ERBB2}), and chromosomal translocations (the BCR-ABL fusion in
chronic myeloid leukaemia). \engterm{Tumour-suppressor genes} are
normal restraints on proliferation; loss-of-function mutations in
both copies of a tumour suppressor (\emph{TP53}, \emph{RB1},
\emph{APC}, the DNA-mismatch-repair genes) remove the restraint.

The cancer-progression model, due to Vogelstein and others, treats
tumour development as a stepwise accumulation of mutations, each
conferring a selective advantage to the cell carrying it. A typical
solid tumour at diagnosis carries $10$ to $100$ \engterm{driver}
mutations and a much larger number of \engterm{passenger} mutations
that confer no selective advantage but accumulated by association.
Recent large-scale sequencing efforts (TCGA in the USA, ICGC
internationally) have catalogued the mutation patterns of dozens of
cancer types and identified hundreds of driver genes. The
engineering reading is that cancer is a failure of the cell's
proliferation-control system caused by mutational degradation of the
control inputs; the therapeutic strategies (targeted inhibition of
activated oncoproteins, immune checkpoint blockade to restore immune
surveillance, restoration of damaged tumour-suppressor function) all
aim to compensate for the lost regulatory inputs or to exploit the
damaged-control state.

\engterm{Mismatch-repair failure and the mutator phenotype}. Cells
with defective mismatch-repair machinery (Lynch syndrome,
microsatellite-unstable colorectal cancer) have mutation rates
$10$ to $1000$ times the normal rate. The accelerated mutation rate
drives faster tumour progression in some contexts but also produces
many neoantigens (mutant proteins recognised by the immune system),
making microsatellite-unstable tumours unusually responsive to
immune checkpoint inhibitors. The mutator phenotype is an instance
of the engineering principle that loss of a single error-correction
stage produces a measurable increase in the overall error rate, with
clinical consequences proportional to the importance of the lost
stage.

\engterm{Hereditary disease}. Inherited diseases caused by mutations
in single genes (monogenic disease) follow Mendelian inheritance
patterns: autosomal dominant (\emph{Huntington's} disease,
\emph{Marfan} syndrome), autosomal recessive (cystic fibrosis,
sickle cell disease), X-linked (haemophilia, Duchenne muscular
dystrophy), or mitochondrial (Leber's hereditary optic neuropathy).
The Online Mendelian Inheritance in Man catalogue (OMIM) lists
$\sim 6000$ such diseases at the time of writing. Polygenic diseases
(common forms of diabetes, heart disease, schizophrenia) arise from
combinations of many small-effect variants; whole-genome association
studies have identified thousands of such variants, with each
typically contributing a fraction of a percent to disease risk. The
engineering reading is that the cell's information-processing system
can fail at many distinct loci, with the failure mode at each locus
determined by the gene's normal function.

\engterm{Replication errors during ageing}. The cells of older
animals carry more mutations than the cells of younger animals.
Mutations accumulate at a rate that integrates the per-division
error rate, the cell-division frequency, and the contribution from
unrepaired DNA damage. The accumulation is implicated in many of the
chronic diseases of ageing (cancer, neurodegeneration,
cardiovascular disease) and is one of the volume's running themes
on cellular failure. Chapter~\ref{vol06:ch07} returns to ageing in
the context of immune senescence; the present chapter limits itself
to the genetic-information side.

\begin{archetype}[Information, failure, and scaling]
The cell's information system exhibits three of the volume's
archetypes in their canonical biological forms. The
\engterm{information archetype} is the digital storage and
controlled retrieval of sequence information at the molecular level,
with error rates corresponding to those of engineered storage. The
\engterm{failure archetype} is the cascade of failure modes
(replication error, repair failure, regulatory loss, proliferation
loss) that produce cancer, hereditary disease, and ageing-associated
pathology. The \engterm{scaling archetype} is the way the same
machinery, sized differently, supports genomes from $4 \times 10^3$
base pairs (small viroids) to $10^{11}$ base pairs (large plant
genomes) with similar per-base error rates and similar architectural
features.
\end{archetype}

\begin{estimation}
\textbf{Question.} Estimate the time it takes to read an entire human
genome on a modern short-read sequencer.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The human haploid genome is $3.2 \times 10^9\,\text{bp}$. To obtain
reliable variant calls, $30\times$ coverage is the working standard:
each base is sequenced an average of $30$ times in independent reads.
The total sequenced bases are $9.6 \times 10^{10}\,\text{bp}$.

A modern Illumina NovaSeq X (current as of 2024) produces approximately
$1.6 \times 10^{13}\,\text{bp}$ of output per $48\,\text{hour}$ run.
The rate is $1.6 \times 10^{13} / (48 \times 3600) = 9.3 \times 10^7\,
\text{bp}/\text{s}$.

The time to sequence one $30\times$ human genome:
\[
T = \frac{9.6 \times 10^{10}}{9.3 \times 10^7} \approx 1030\,\text{s} \approx 17\,\text{minutes}.
\]

In practice the instrument multiplexes many samples per run (the
$48\,\text{h}$ run produces approximately $160$ human genomes at
$30\times$ each), so the per-sample wall-clock time is a fraction of
the run time but the throughput per sample is the figure above.

\smallskip
\textit{Postmortem.} The estimate is the throughput limit, not the
end-to-end time. Sample preparation, library construction, quality
control, base calling, alignment, and variant calling each add
time. The actual turn-around for a clinical genome sequencing is
typically $1$ to $2\,\text{weeks}$ at the time of writing, dominated
by sample handling rather than by the instrument itself. The
sequencing throughput has reached the point where it is no longer
the rate-limiting step; the rate-limiting step is now the
informatics, the interpretation, and the clinical reporting. The
shift identifies the next engineering bottleneck and the focus of
Chapter~\ref{vol06:ch11}.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Simulate genetic drift in a small population]
The reader simulates the Wright-Fisher model of genetic drift in a
finite population, observes the time-to-fixation distribution, and
compares to the analytic predictions.

\textbf{Track}. Simulation.
\textbf{Hazard class}: none.

\textbf{Model}. A diploid population of $N$ individuals, with a single
locus carrying two alleles ($A$ and $a$). Each generation, $2N$
gametes are sampled with replacement from the previous generation's
allele pool. Initial allele frequency $p_0 = 0.5$.

\textbf{Procedure}.

\begin{enumerate}[leftmargin=*]
\item Implement the Wright-Fisher sampling step in a programming
  language of your choice.
\item For population sizes $N \in \{10, 100, 1000\}$, simulate
  $1000$ independent populations and record the generation at which
  each population fixes one allele or the other.
\item Plot the empirical distribution of fixation times and compute
  the mean.
\item Compare with the analytic prediction that the expected time to
  fixation conditional on $p_0 = 0.5$ scales as $\sim 2.8 N$ generations.
\item Repeat with a selective advantage $s = 0.01$ for the $A$ allele
  and report how the probability of $A$ fixation depends on $N s$.
\end{enumerate}

\textbf{Deliverable}. A five- to ten-page report with one
fixation-time-distribution figure, one fixation-probability-vs-$N s$
figure, source code, and a paragraph identifying which analytic
prediction your simulation reproduces most accurately and which least.

\textbf{Time}. Four to eight hours.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
The human haploid genome is $3.2 \times 10^9\,\text{bp}$. Compute
the information content in bits and in bytes. Compare to the storage
capacity of a typical USB flash drive (current as of 2024).
\end{exercise}

\begin{exercise}
DNA polymerase has a base-incorporation rate of approximately
$1000\,\text{bp}/\text{s}$ in \emph{E.~coli}. Compute the time
required to replicate the entire \emph{E.~coli} genome
($4.6 \times 10^6\,\text{bp}$) from a single bidirectional origin.
Compare to the observed doubling time of $30\,\text{min}$.
\end{exercise}

\begin{exercise}
The persistence length of DNA is $L_p = 50\,\text{nm}$. Compute the
number of base pairs in one persistence length and the bending energy
required to curve a DNA segment of length $L_p$ into a complete
circle ($U = 2\pi^2 k_B T L_p/L$). Use $T = 310\,\si{\kelvin}$.
\end{exercise}

\begin{exercise}
Compute the energetic cost (in ATP equivalents) of synthesising a
mRNA of $1000\,\text{nucleotides}$ given the cost per nucleotide of
$2$ ATP equivalents (one for the triphosphate bond, one for the
nucleotide diphosphate synthesis). Compute the cost of translating
the $333$-amino-acid protein it encodes.
\end{exercise}

\begin{exercise}
The mutation rate in \emph{E.~coli} is approximately $10^{-9}$ per
base pair per generation. Compute the expected number of mutations
per cell per generation in the $4.6 \times 10^6\,\text{bp}$ genome.
Compute the expected number of mutations per gene per generation for
a typical $1000\,\text{bp}$ gene.
\end{exercise}

\begin{exercise}
The human germline mutation rate is approximately
$1.2 \times 10^{-8}$ per base pair per generation. Compute the
expected number of new mutations in a child relative to the parents
in the $6.4 \times 10^9\,\text{bp}$ diploid genome.
\end{exercise}

\begin{exercise}
A polymerase has an intrinsic error rate of $10^{-5}$ from base
selection and proofreading reduces it to $10^{-7}$. Mismatch repair
catches $99\,\%$ of the remaining errors. Compute the effective
error rate at fixation.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the expected time to fixation by genetic drift in a haploid
Wright-Fisher population of size $N$, starting from allele frequency
$p_0$, conditional on eventual fixation of the focal allele.
\end{exercise}

\begin{exercise}
Derive the Hardy-Weinberg equilibrium for diploid allele frequencies
under random mating and no selection, mutation, or migration.
\end{exercise}

\begin{exercise}
Derive the expected number of synonymous and non-synonymous
substitutions per codon under a uniform random-codon-substitution
model, given the genetic-code table. Use the result to motivate the
$d_N/d_S$ statistic used in molecular-evolution studies.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the total mass of DNA in the human body, given that there
are approximately $4 \times 10^{13}$ cells each containing
$\sim 6\,\text{pg}$ of DNA.
\end{exercise}

\begin{exercise}
Estimate the rate at which a typical cell repairs oxidative DNA
damage, given that approximately $10\,000$ oxidative lesions occur
per cell per day and a healthy cell maintains a steady-state count
of fewer than $100$ unrepaired lesions.
\end{exercise}

\begin{exercise}
Estimate the number of distinct human protein-coding gene sequences
that are perfectly preserved across all $8 \times 10^9$ people alive
today (current as of 2024). Use the per-base mutation rate and the
expected accumulated mutation count per gene.
\end{exercise}

\begin{exercise}
Estimate the cost (in dollars per base pair) of writing a designed
DNA sequence by current commercial gene-synthesis services (current
as of 2024). Compare to the cost per base pair of reading by
short-read sequencing.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Simulate the accumulation of point mutations in a single
$3 \times 10^9\,\text{bp}$ genome over $10^4$ generations at the
mammalian mutation rate. Plot the mutation count per gene over time
and identify the variance.
\end{exercise}

\begin{exercise}
Implement a simulation of the Luria-Delbr\"uck fluctuation test for a
bacterial population. Show that the variance of mutant counts across
parallel cultures distinguishes random spontaneous mutation
(Luria-Delbr\"uck distribution) from induced mutation (Poisson
distribution).
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design an error-correcting DNA storage system that encodes
$1\,\text{MB}$ of arbitrary digital information into a synthetic DNA
sequence. Specify the encoding scheme, the redundancy level, and the
expected retention time under refrigerated storage. Compare to
current commercial DNA-storage demonstrations.
\end{exercise}

\begin{exercise}
Design a sequencing strategy for the genome of a previously
unsequenced eukaryote of $10^9\,\text{bp}$. Specify the mix of
short-read and long-read sequencing, the depth, and the expected
cost (current as of 2024 prices). Identify which steps of the
assembly are likely to be hardest.
\end{exercise}

\begin{exercise}
Design a CRISPR-Cas9 guide RNA for editing a specific position in the
human genome. Specify the constraints on guide RNA design (length,
target specificity, PAM sequence) and the off-target risks to be
checked.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A patient's tumour exhibits microsatellite instability and an
unusually high mutation rate. Identify the likely failed DNA-repair
pathway and propose a clinical confirmation.
\end{exercise}

\begin{exercise}
A bacterial strain in industrial fermentation drifts in its
production yield over many batches. Identify three possible
information-level causes (plasmid instability, mutation in the
production gene, regulatory drift, host evolution) and propose a
diagnostic measurement for each.
\end{exercise}

\begin{exercise}
A child's whole-genome sequencing identifies $50$ new mutations
relative to the parents, of which two are in protein-coding regions.
Estimate the probability that one of the two coding-region mutations
is pathogenic and propose a follow-up workflow.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A patient with xeroderma pigmentosum develops severe skin cancers
from minimal UV exposure. Identify the failed repair pathway
(nucleotide excision repair) and trace the chain from defective
repair to the observed cancer phenotype.
\end{exercise}

\begin{exercise}
A new biopharmaceutical produced in mammalian cell culture shows
unexpected sequence variants at $0.5\,\%$ frequency in mass-spectrometric
analysis of the product. Identify three possible sources (replication
error, transcription error, translation error, post-translational
modification) and propose a measurement that would distinguish them.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A direct-to-consumer genetic-testing company reports a customer's
risk of a complex disease based on a polygenic-risk score. Identify
two engineering checks the customer should apply to the reported
probability (population calibration, ancestry-specific accuracy,
confidence interval) and write a one-paragraph reply to the customer's
question ``does this mean I will get the disease?''.
\end{exercise}

\begin{exercise}
A press release describes a ``DNA-based diagnostic'' that detects
disease ``before any symptoms appear'' with $99.9\,\%$ accuracy.
Identify the engineering questions (sensitivity, specificity, base
rate, positive predictive value) the accuracy claim does not answer
and write a one-paragraph reply.
\end{exercise}

\begin{exercise}
A textbook describes the genome as a ``blueprint'' for the organism.
Identify two ways in which the blueprint metaphor is accurate
(sequence-defined inheritance, deterministic transcription rules) and
two ways in which it is misleading (the genome does not specify a
unique adult organism, and the regulatory layer is at least as
important as the coding sequence).
\end{exercise}

\begin{exercise}
A vendor advertises a ``synthetic genome'' that can be installed in
an empty cellular chassis to produce a fully functional cell.
Identify the engineering constraints the design must satisfy
(replication-origin compatibility, codon usage matched to host tRNA
pool, essential-gene coverage, regulatory completeness) and write a
one-paragraph reply identifying which constraint is most likely
underspecified in the advertisement.
\end{exercise}

\begin{exercise}
A press release claims that a new gene-editing therapy ``corrects''
a disease mutation in $99\,\%$ of patient cells. Identify the
engineering questions the percentage does not answer (which tissues
were edited, what fraction of stem cells were edited, what off-target
edits occurred, whether the correction is heritable) and write a
one-paragraph reply.
\end{exercise}

\begin{exercise}
A reader argues that the discovery of the genetic code in the
$1960$s exhausted the engineering content of molecular biology.
Write a one-paragraph reply naming two layers of regulatory
information (chromatin state, non-coding RNA networks) that the
code does not capture and that have emerged as engineering targets
in the intervening decades.
\end{exercise}
