\chapter{Engineering project management}
\label{vol09:ch10}

\epigraph{The project is a temporary system; the discipline is making it a system at all.}{}

\chapmeta{Half-life: medium-life. Archetypes: scaling, control, balance. Exercise target: 25. Project track: physical.}

\noindent
Engineering project management is the discipline of organising people, resources, time, and risk to deliver an engineered artefact or capability within budget and schedule. The discipline differs from operations management (which addresses ongoing production) in its temporary nature: every project has a start, an end, and a defined deliverable. The discipline differs from research (which prioritises discovery) in its constraint-driven nature: the project must deliver against the specification. Engineering project management is part-and-parcel of engineering itself; an engineer who can design but cannot deliver is incomplete. This chapter develops the engineer's working understanding of projects as systems, work-breakdown structure and estimation, scheduling methods (PERT, CPM, critical chain), risk management, stakeholder analysis, earned-value management, programme and portfolio management, and the characteristic failure modes of engineering projects.

\section{Projects as systems}\pathtag{standard}

A project is a temporary organisation created to produce a unique deliverable. The temporary feature distinguishes projects from ongoing operations; the unique feature distinguishes projects from batch production. The combination drives the project's management discipline.

The project's stakeholders form its environment: the customer (who pays), the sponsor (who is accountable), the team (who executes), the suppliers (who provide), the regulators (who certify), the users (who operate), and others affected by the project's outcome. The stakeholder map is the first artifact of any non-trivial project; misunderstanding the stakeholders is the most common source of project failure.

The project's triangle is cost, time, and scope (sometimes drawn as a triangle with quality at the centre). The three are coupled: increasing scope requires more time or more cost; cutting cost forces compromises in scope or schedule. The project manager's job is to maintain explicit agreement on the trade-offs as conditions change.

A project's life cycle has phases: initiation (definition and approval), planning (work-breakdown, estimation, scheduling, risk), execution (the actual work), monitoring and control (tracking against plan, addressing variances), and closeout (handover, documentation, retrospective). Standard methodologies (PMBOK, PRINCE2, agile family) organise the work; the choice of methodology depends on the project's nature.

Engineering projects sit in a spectrum: from small (an internal tool, a single experiment) to large (a satellite, a power plant, a refinery) to mega (a nuclear plant, a national high-speed-rail line, a space programme). The scaling of project management is non-linear: a mega-project is not just $1000$ times a small project; the dynamics, the failure modes, the management methods are different.

\begin{archetype}[coupling: project as temporary system]
A project is a system whose components are people, machines, deliverables, milestones, and risks, coupled by the project plan. The plan is the central abstraction: it expresses the work, the dependencies, the resources, the timeline. The plan changes as the project unfolds; managing the plan is managing the project. The engineer who treats the plan as a static document at project start has misunderstood projects; the plan is a living artefact that evolves with the project.
\end{archetype}

\section{Work-breakdown structure and estimation}\pathtag{core}

The work-breakdown structure (WBS) is a hierarchical decomposition of the project's work into manageable elements (work packages). The decomposition is by deliverable, not by activity: each level of the WBS represents a deliverable or sub-deliverable; the leaves are work packages that can be estimated and assigned.

A well-formed WBS has the $100\%$ rule: the total work at each level is the sum of the work below it; no work is omitted, no work is double-counted. The hierarchy depth is moderate ($3$ to $7$ levels for most projects); the work packages are sized so that they can be planned, executed, and tracked.

Estimation is the assignment of time and cost to work packages. Standard estimation methods:

\textbf{Analogous:} compare to past similar projects, adjust for differences. Fast, broad, low accuracy.

\textbf{Parametric:} use models that relate scope to effort (e.g., function points for software, lines-of-code, COCOMO; building cost per square foot; oil-platform cost per barrel of capacity). Moderate effort, moderate accuracy.

\textbf{Bottom-up:} decompose each work package, estimate each piece, sum up. High effort, high accuracy for well-understood work.

\textbf{Three-point (PERT):} estimate optimistic, most likely, pessimistic; combine as $(O + 4M + P)/6$ with variance $((P - O)/6)^2$. Captures uncertainty; useful for new or risky work.

Estimation has systematic biases. Optimism bias: estimators systematically underestimate effort and overestimate productivity. Anchoring: initial estimates anchor subsequent revisions. Conjunction fallacy: estimating the joint duration is dominated by reasoning about the activities, ignoring the way uncertainties combine. Mitigation: use reference-class forecasting (compare to similar past projects' actual durations), explicit uncertainty intervals, and post-project retrospection to calibrate.

\section{Scheduling: PERT, CPM, critical chain, resource-constrained}\pathtag{core}

The project schedule expresses the time dimension: when each activity starts and ends, what depends on what, what is critical.

The critical path method (CPM) takes activity durations and precedence relations, computes the earliest start and finish, latest start and finish, total float (slack), and identifies the critical path (activities with zero float). The critical path is the chain that determines the project duration; delays to critical-path activities delay the project; delays to non-critical activities consume float without affecting completion.

PERT (Program Evaluation and Review Technique) extends CPM with the three-point estimation for activity durations, producing a probability distribution over project completion time. PERT's mathematical assumption (each activity duration is approximately beta-distributed; the critical path is fixed) is a simplification; Monte Carlo simulation gives a more accurate picture for projects with significant uncertainty.

Critical chain project management (CCPM, Goldratt 1997) emphasises resource constraints and explicit buffers. The critical chain is the longest path through the project considering both precedence and resource constraints; buffers are placed at the chain's end (project buffer) and at the convergence of feeding chains (feeding buffer). The method addresses the issues that classical CPM ignores (resource conflict, student-syndrome procrastination, Parkinson's law expansion of work).

Resource-constrained project scheduling (RCPSP) is the MILP formulation of the same problem: minimise duration subject to precedence and resource constraints. Solvers handle moderate problems; commercial project-management software (Primavera, Microsoft Project) uses heuristics with optional optimisation.

Modern projects use schedule management as a continuous discipline: weekly or daily updates, replanning when variances exceed thresholds, schedule-risk analysis before milestone commitments. The schedule is a planning tool, a tracking tool, and a communication tool; treating it as just one of the three weakens the project's control.

\section{Risk management: registers, qualitative and quantitative analysis}\pathtag{standard}

Risk management is the systematic identification, analysis, prioritisation, and treatment of risks. A risk is a potential event that, if it occurred, would affect the project's objectives.

The risk register is the central artefact: a list of identified risks with their attributes (description, probability, impact, response, owner, status). Risks are added as identified; closed as they materialise or pass; reviewed periodically for updates.

Qualitative risk analysis assigns probability and impact scores (typically $1$ to $5$ scales), computes a risk priority score (probability $\times$ impact), and ranks risks for attention. High-priority risks get explicit treatment plans; low-priority risks are documented and monitored.

Quantitative risk analysis assigns probability distributions and impact distributions, runs Monte Carlo simulation, and produces probabilistic forecasts (P50, P80, P90 cost and schedule). The quantitative analysis is more rigorous and more useful for large projects; the qualitative analysis is faster and adequate for smaller projects.

Risk responses include: avoid (eliminate the cause), transfer (insurance, contracts), mitigate (reduce probability or impact), accept (acknowledge and reserve contingency). The response choice depends on the risk's characteristics and the cost of each response option.

Megaprojects have characteristic risks: cost overruns ($28\%$ average for transport mega-projects, per Flyvbjerg's research), schedule delays, scope changes, stakeholder conflicts, technology immaturity, political shifts, regulatory delays. The risk register for a mega-project is itself a substantial document; the analysis informs go/no-go decisions and reserve estimates.

\section{Stakeholder analysis and communication plans}\pathtag{standard}

The stakeholder analysis identifies who has interest in or influence over the project, characterises their concerns and expectations, and informs the communication and engagement strategy. The standard outputs include the stakeholder register, the power-interest matrix (high-power-high-interest stakeholders get the most attention), and the engagement plan.

The communication plan specifies what is communicated, to whom, how often, in what form. Standard elements: project status reports (frequency, audience, content), milestone reviews, risk reports, change reports, decision logs. The communication burden is significant on large projects; under-communication produces stakeholder surprise; over-communication produces stakeholder fatigue.

Stakeholder analysis includes recognising the conflicts: different stakeholders have different objectives; the project manager facilitates the negotiation that produces a project consistent with the major stakeholders' interests. Some conflicts cannot be resolved within the project's scope; they must be escalated, deferred, or accepted as risk.

Effective stakeholder management is a soft skill on top of the hard discipline. The engineer who is fluent in stakeholder management can lead larger projects; the engineer who cannot is restricted to projects where someone else handles the stakeholders.

\section{Earned-value management}\pathtag{standard}

Earned-value management (EVM) is a quantitative method for tracking project performance against the plan. The standard metrics:

\textbf{Planned value (PV):} the budgeted cost of work scheduled at the current date.
\textbf{Earned value (EV):} the budgeted cost of work actually completed at the current date.
\textbf{Actual cost (AC):} the actual cost incurred at the current date.

From these:

\textbf{Schedule variance (SV):} $\text{EV} - \text{PV}$. Positive = ahead of schedule; negative = behind.

\textbf{Cost variance (CV):} $\text{EV} - \text{AC}$. Positive = under budget; negative = over.

\textbf{Schedule performance index (SPI):} $\text{EV}/\text{PV}$. $> 1$ = ahead; $< 1$ = behind.

\textbf{Cost performance index (CPI):} $\text{EV}/\text{AC}$. $> 1$ = under budget; $< 1$ = over.

\textbf{Estimate at completion (EAC):} forecast of total cost. The simplest formula: $\text{EAC} = \text{Budget at Completion} / \text{CPI}$, assuming the CPI continues. Other formulas use SPI or weighted combinations.

EVM is the standard tool for projects with significant cost and schedule discipline (government contracts, defence acquisition, large engineering projects). The discipline requires baselined plans, accurate progress measurement, and disciplined data entry; the payoff is early warning of cost and schedule problems that intuitive tracking misses.

EVM limitations include: dependence on accurate progress measurement (over-reporting "complete" leads to optimistic EVM); insensitivity to scope quality (a $90\%$-complete deliverable may not be useful); challenges with agile or research-style work where the WBS is iterative.

\begin{estimation}
\textbf{Question.} Estimate the cost performance index (CPI) below which a project is in serious trouble and likely to fail.

\textbf{Working.} Industry rules of thumb suggest that projects with CPI $< 0.90$ at the $20\%$ completion point rarely recover. The reasoning: the trajectory of CPI tends to stay roughly constant or worsen; an early CPI of $0.85$ implies a total cost overrun of $\sim 18\%$, which exceeds typical contingency. The threshold varies by industry: aerospace and defence projects routinely operate at CPI $< 1.0$ with managed overrun; commercial construction and software projects have less tolerance. The engineering response to low CPI: identify the root causes (estimation error, scope growth, productivity loss, rework), implement specific corrections, re-baseline if the corrections do not restore performance.
\end{estimation}

\section{Programme and portfolio management}\pathtag{mastery}

A programme is a group of related projects managed in a coordinated way to achieve benefits unavailable from the projects individually. A portfolio is a collection of programmes and projects grouped for management efficiency to meet strategic objectives.

Programme management coordinates dependent projects: sharing resources, aligning schedules, managing inter-project risks, integrating deliverables. Standard examples: aerospace programmes (launcher development, payload development, mission operations); large infrastructure (a city's transit system across multiple lines); product families (a product platform with variants for different markets).

Portfolio management selects and balances projects against strategic objectives: which projects to fund, when to start, how to allocate resources across the portfolio. Standard methods: scoring models, portfolio optimisation, balanced scorecards. The discipline includes regular portfolio review and reallocation as conditions change.

Programme and portfolio management add a layer of complexity over single-project management. The methodologies (Managing Successful Programmes, Standard for Program Management) provide structure but do not eliminate the inherent complexity. Megaprojects often run as programmes; very large public projects (e.g., the European Space Agency's launcher programmes, US defence acquisition programmes) include both programme and portfolio layers.

\section{Failure: schedule cynicism, scope creep, hidden dependencies, megaproject pathologies}\pathtag{core}

Engineering projects fail in characteristic ways.

\textbf{Schedule cynicism:} the team or organisation no longer believes the schedule; everyone plans to their own private timeline, ignoring the published schedule. Symptoms: routinely missed milestones, vague schedule status reports, finger-pointing when delays are exposed. Mitigation: realistic estimation; transparent tracking; consequence for the schedule (decisions that depend on schedule status); explicit acknowledgement when the schedule is wrong and re-baselining.

\textbf{Scope creep:} the project's scope expands over its life without commensurate adjustment of cost and schedule. The expansion is small at each step but cumulative; the project ends up significantly over budget and behind schedule. Mitigation: change-control discipline; explicit cost-and-schedule impact analysis for every scope change; stakeholder approval; periodic baseline review.

\textbf{Hidden dependencies:} the schedule reflects the activities the team understands but omits the dependencies the team does not. Examples: a critical supplier whose lead time was not in the schedule; an approval that requires regulatory action with its own timeline; a key team member whose availability was not constrained. The hidden dependency surfaces as a surprise delay. Mitigation: thorough WBS and schedule review; supplier and approval lead-time tracking; explicit cross-team dependency mapping.

\textbf{Megaproject pathologies:} large projects suffer characteristic failures. Optimism bias produces optimistic initial estimates. Strategic misrepresentation (motivated under-estimation to secure approval) compounds the optimism. Lock-in (sunk cost, momentum) prevents cancellation of failing projects. Scope expansion as politically motivated additions accumulate. The pathologies are documented in Flyvbjerg's work; the average cost overrun for transport mega-projects is $28\%$; for rail, $45\%$; for IT mega-projects, $107\%$.

\textbf{Communication failures:} the stakeholders are misaligned, the team is confused, the deliverables do not match expectations. The project is technically successful but socially unsuccessful. Mitigation: stakeholder management; clear communication; explicit acceptance criteria.

\textbf{Risk-management theatre:} the project maintains a risk register but does not use it to drive decisions. Risks are documented and ignored; when they materialise, no response was prepared. Mitigation: tie risk management to decision-making; review the register at every major milestone; require risk owner accountability.

\textbf{EVM data corruption:} EVM reports paint a rosy picture; reality is worse. Causes: over-reporting "complete"; misallocation of cost; rebaseline that hides past variance. Mitigation: independent EVM audits; objective progress measures; transparency in data entry.

\textbf{Cultural mismatch:} the project team's methods do not match the organisation's culture or the customer's expectations. A government project executed by a startup-style team; an agile-team approach imposed on a regulated industry; a discipline gap between technical and managerial staff. Mitigation: methodology choice that matches the context; explicit cultural translation; bridge roles between team and stakeholders.

The 1986 Challenger disaster (\cite{acc:nasa-challenger}) had project-management roots: schedule pressure from political commitments; cost pressure from austerity; communication failure between engineers and managers about O-ring temperature limits. The technical fault (O-ring failure at low temperature) was known and discussed; the management decision-making process did not act on the warnings. The Rogers Commission's findings include both technical and managerial recommendations.

\begin{principle}[The estimate-and-track principle]
The project's primary discipline is estimation and tracking. The estimate at start is the engineering assertion of what the project will cost and take. The tracking through execution is the engineering check on whether the estimate is holding. The variance between estimate and actual is the engineering signal of misunderstanding (the estimate was wrong) or trouble (the execution is not going as planned). A project without estimation cannot be tracked; a project without tracking is not managed. The discipline is the same at every project scale; only the formality varies.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Lead a six-week engineering mini-project end-to-end]
\textbf{Objective.} Lead an engineering mini-project of approximately six weeks' duration end-to-end with explicit project management discipline. Use earned-value management to track performance. Produce a closeout report.

\textbf{Method.} Choose a project: a software tool you intend to use; a hardware project you intend to complete; a deliverable for a course or internship; a personal engineering project. Develop a WBS, estimate effort, schedule activities, identify risks, establish a baseline. Track weekly: planned value, earned value, actual cost. Compute SPI, CPI, and forecast EAC at each tracking point. Apply change control when scope changes. At completion, produce a closeout report: baseline vs actual cost and schedule, variances and root causes, lessons learned, recommendations for future projects.

\textbf{Deliverables.} The WBS; the baseline schedule and budget; the weekly tracking data; the closeout report.

\textbf{Hazard.} The methodology is fictional in nature compared to a "real" project; the value is in practicing the discipline. The deliverable is intended to be useful even if imperfect.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A project has activities $A$ ($5\,\text{d}$), $B$ ($3\,\text{d}$, depends on $A$), $C$ ($4\,\text{d}$, depends on $A$), $D$ ($2\,\text{d}$, depends on $B$ and $C$). Compute the critical path and project duration.
\end{exercise}

\begin{exercise}
A PERT activity has optimistic $4\,\text{d}$, most likely $7\,\text{d}$, pessimistic $13\,\text{d}$. Compute the expected duration and standard deviation.
\end{exercise}

\begin{exercise}
At month $3$ of a project, $\text{PV} = \$300{,}000$, $\text{EV} = \$280{,}000$, $\text{AC} = \$310{,}000$. Compute SV, CV, SPI, CPI.
\end{exercise}

\begin{exercise}
For the previous, compute the EAC assuming the CPI continues. Total budget is $\$1{,}000{,}000$.
\end{exercise}

\begin{exercise}
A risk has probability $0.3$ and impact $\$500{,}000$. Compute the expected risk cost.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the earliest-start and latest-start times in CPM from the activity precedence graph.
\end{exercise}

\begin{exercise}
Derive the PERT formula for activity duration mean and variance from the beta-distribution assumption.
\end{exercise}

\begin{exercise}
Derive the EAC formula assuming constant CPI from the relationship between BAC, EV, and AC.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the contingency reserve required for a project with $20$ identified risks and approximate $\$200{,}000$ expected risk cost; choose a confidence level.
\end{exercise}

\begin{exercise}
Estimate the schedule overrun probability for a software project with three-point estimates summing to $90\,\text{d}$ at the most-likely value and $130\,\text{d}$ at the pessimistic.
\end{exercise}

\begin{exercise}
Estimate the staffing required for a $12$-month software-development project with $4000\,\text{h}$ total effort and account for $80\%$ productivity.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate a Monte Carlo PERT analysis: generate activity durations from triangular distributions; report the P50 and P80 project durations.
\end{exercise}

\begin{exercise}
Simulate the effect of buffer placement (critical-chain vs traditional CPM) on a small project network with stochastic durations.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design the WBS for an engineering mini-project of your choice; estimate each work package; identify the critical path.
\end{exercise}

\begin{exercise}
Design a risk register for an aerospace project: ten major risks, probability, impact, response, owner.
\end{exercise}

\begin{exercise}
Design a stakeholder communication plan for a public-infrastructure project: stakeholder map, communication frequency, format.
\end{exercise}

\begin{exercise}
Design an EVM tracking process for a six-month engineering project: data collection, reporting, decision triggers.
\end{exercise}

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A team's project has SPI $0.85$ and CPI $0.80$ at the $30\%$ completion point. State the engineering response.
\end{exercise}

\begin{exercise}
A team's project has accumulated significant scope creep over six months; the customer requests one more change. State the engineering response.
\end{exercise}

\begin{exercise}
A team is choosing between PERT and Monte Carlo for schedule risk analysis on a moderate-uncertainty project. State the engineering trade-offs.
\end{exercise}

\begin{exercise}
A team's project is on track per EVM but the deliverable quality is degrading. State the engineering response.
\end{exercise}

\begin{exercise}
A team is launching a mega-project with $\$1B$ budget and $7$-year timeline. State the engineering considerations for risk management, given typical megaproject pathologies.
\end{exercise}

\begin{exercise}
A team's project has consistent schedule slippage; the team requests "more aggressive scheduling." State the engineering response.
\end{exercise}

\begin{exercise}
A team's project has been baselined; new information suggests the baseline is significantly wrong. State the engineering response.
\end{exercise}

\begin{exercise}
A team is choosing between agile and waterfall for an engineering project with regulatory deliverables and an unhappy customer who has complained about communication. State the engineering trade-offs and the role of stakeholder analysis in the methodology choice.
\end{exercise}
