\chapter{Failure as engineering subject}
\label{vol10:ch01}

\epigraph{Things fail. The discipline is what we do about it.}{}

\chapmeta{Half-life: durable. Archetypes: failure, coupling. Exercise target: 20. Project track: physical.}

\noindent
Engineering failure is unavoidable. Components wear, materials degrade, software bugs surface, operators err, organisations misjudge, regulators miscalibrate. The unavoidability is not a defect of the discipline; it is its central condition. The engineer who recognises failure as the condition of practice and not the negation of it builds systems that fail well: predictably, recoverably, with learning. This chapter, the opener of Volume X, develops failure as a first-class engineering subject: why it deserves a volume, the taxonomy of failure, the standard frameworks (Swiss cheese, high-reliability organisations, resilience engineering), the disposition of treating failure as data rather than shame, the reader's posture for productive failure study, and the meta-failure of failure analysis itself.

\section{Why failure deserves a volume}\pathtag{standard}

The engineering literature is dominated by success: how to build, how to optimise, how to control. The failure literature exists but is fragmented across domains: aerospace accident reports, civil engineering forensic engineering, software post-mortems, chemical process safety analyses, medical-device safety records, nuclear-plant probabilistic risk assessments. The fragmentation hides the patterns; the patterns recur across domains because the underlying mechanisms (component degradation, coupling, common-mode failure, human-organisational factors) recur.

Volume X is the engineer's failure-mode reference, organised to support cross-domain transfer. The Therac-25 software-radiation-dose accident lives in medical-device literature; the mechanism (operator-input race condition in a state machine with insufficient interlocks) is a general lesson for any cyber-physical system. The Tacoma Narrows torsional flutter lives in bridge literature; the mechanism (aeroelastic coupling at a frequency not directly excited) recurs in aircraft wings, building wind response, electrical-grid sub-synchronous oscillation. The 1986 Challenger O-ring blow-by lives in space-program literature; the mechanism (cold-weather sealing margin reduction, with engineering warnings overridden by schedule pressure) recurs in every safety-critical organisation.

The unifying thesis: engineering is the discipline by which measured reality becomes reliable intervention under constraint, failure, scale, and responsibility (the book's thesis, restated). Failure is one of the four constraints; the volume is failure's full treatment.

The economic case is also strong. The cost of failures dominates many engineering operating budgets: aerospace maintenance and reliability investments, medical-device regulatory compliance, financial-systems uptime engineering, infrastructure deferred-maintenance backlogs. The investment in failure prevention, detection, mitigation, and recovery is engineering work; it is not overhead, it is the work.

\begin{archetype}[failure: as engineering's first-class subject]
Failure is not the absence of success; it is a category of engineering phenomenon with its own mechanisms, statistics, organisational dynamics, and discipline. The engineer treats failure with the same rigour as design: characterise the mechanism, quantify the rate, identify the precursors, design the mitigation, monitor for the precursors, respond to the events, learn from each. The half-discipline (study only success) produces engineers who are surprised by failures; the full discipline (study both) produces engineers who anticipate failures and engineer around them.
\end{archetype}

\section{The taxonomy of failure}\pathtag{core}

Failures classify along multiple axes. The standard taxonomies:

\textbf{By mechanism:} mechanical (fatigue, fracture, wear, corrosion, plastic deformation), thermal (overheating, thermal cycling, thermal runaway), chemical (corrosion, attack, contamination), electrical (overvoltage, overcurrent, dielectric breakdown), software (bugs, race conditions, data corruption), human (error, slip, mistake, violation), organisational (poor communication, schedule pressure, normalisation of deviance), environmental (overload from external conditions), and combinations.

\textbf{By rate:} sudden (catastrophic, single event) vs gradual (degradation, accumulation, drift). Sudden failures often have long latent periods; gradual failures often appear sudden when a threshold is crossed.

\textbf{By cause:} random (probabilistic, no specific reason for this event vs that one) vs systematic (a specific cause that affects all instances). Random failures are managed by reliability theory; systematic failures require root-cause analysis and design changes.

\textbf{By scope:} component (one part fails) vs subsystem (multiple parts fail together) vs system (the entire system loses function) vs system-of-systems (multiple systems lose function together). The scope determines the appropriate response and the consequence severity.

\textbf{By consequence:} negligible (no operational impact), minor (recoverable with normal procedures), major (recovery requires special procedures), catastrophic (recovery impossible, lives lost or environment damaged), existential (organisation-ending or society-affecting). The consequence classification informs the investment in prevention and mitigation.

\textbf{By detectability:} immediate (the failure is obvious as it happens) vs latent (the failure exists but is not detected for some time) vs hidden (the failure is never detected, only inferred from consequences). Latent and hidden failures are particularly dangerous; they accumulate behind the visible failure rate.

\textbf{By predictability:} predictable (the failure was expected at this point in life or under these conditions) vs surprising (the failure occurred outside the expected envelope) vs unprecedented (no prior experience of this failure mode). Predictable failures are managed; surprising failures expose modelling errors; unprecedented failures require new understanding.

The taxonomies are not mutually exclusive; a single failure can be classified along multiple axes. The Bhopal disaster (1984) was chemical (methyl isocyanate reaction), gradual (refrigeration deferral, instrument neglect), systematic (organisational, design, regulatory factors all contributing), system-wide in scope, catastrophic in consequence, latent in many factors, and at the time partially unprecedented in its specific combination.

\section{The Swiss-cheese model and its critics}\pathtag{standard}

James Reason's Swiss-cheese model (1990) is the most-cited framework for systemic failure. The model: an accident requires alignment of holes through multiple defensive layers; each layer (engineering controls, procedures, training, supervision, regulation) has gaps; failures occur when the gaps align. The layers are imperfect; the engineering work is to keep the gaps small and uncorrelated, so that alignment is rare.

The model has practical merit: it directs attention to multiple defensive layers rather than single causes; it acknowledges that any single layer is imperfect; it justifies investment in diverse defences.

Critics of the Swiss-cheese model identify limitations:

\textbf{The model is descriptive, not predictive.} It describes how accidents happen retrospectively but does not predict where the next accident will be.

\textbf{The model treats failures as discrete events.} Many failures are gradual processes; the discrete-event framing misses the slow accident.

\textbf{The model emphasises sequential layers.} Many failures involve nonlinear interactions across multiple layers simultaneously, not a sequential traversal.

\textbf{The model risks blame allocation.} The "holes" in particular layers can be interpreted as the responsibility of the people who maintain that layer; the system-level perspective gets lost.

Alternative frameworks include:

\textbf{High-reliability organisations (HRO):} Weick and Sutcliffe (2007), focused on organisations that operate safely despite high inherent risk (aircraft carriers, nuclear plants, hospital emergency rooms). HRO characteristics: preoccupation with failure, reluctance to simplify, sensitivity to operations, commitment to resilience, deference to expertise. The framework is prescriptive.

\textbf{Resilience engineering:} Hollnagel et al. (2006), focused on the system's ability to maintain function in varied conditions including disruptions. The framework treats safety as an emergent property rather than as the absence of failures.

\textbf{Normal Accidents theory:} Perrow (1984), arguing that some systems are so complex and tightly coupled that accidents are inevitable. The framework is partly fatalistic but provides a useful diagnostic: high complexity plus tight coupling produces inherent accident-proneness.

\textbf{Drift to failure:} Dekker (2011), focused on the gradual erosion of safety margins as organisations adapt to operational pressures. The framework explains many slow accidents better than the Swiss-cheese model.

The engineering use of these frameworks: they inform the diagnostic vocabulary, the questions asked during investigation, and the structure of safety management systems. No single framework is complete; the engineer uses what fits the problem.

\section{The HRO and resilience-engineering traditions}\pathtag{standard}

High-reliability organisations (HRO) study organisations that achieve much lower accident rates than expected for their risk environment. Examples: nuclear-powered aircraft carriers (intense flight operations with very few incidents), nuclear power plants (decades of operations with rare significant events), air-traffic control (millions of operations annually with very few accidents).

The HRO characteristics:

\textbf{Preoccupation with failure:} the organisation actively seeks indications of potential failure rather than waiting for failures to occur. Near-misses are investigated; weak signals are followed up; complacency is recognised as a failure mode in itself.

\textbf{Reluctance to simplify:} the organisation resists the temptation to explain events with simple, single-cause stories. Complex events deserve complex analysis; the simplification often loses the important nuance.

\textbf{Sensitivity to operations:} the organisation pays attention to the front-line workers' experience. The people actually doing the work see things that managers and analysts do not.

\textbf{Commitment to resilience:} the organisation invests in the capacity to recover from disturbances rather than only in their prevention. Recovery capability is as important as prevention capability.

\textbf{Deference to expertise:} authority migrates to the people with the relevant expertise during high-tempo or high-risk operations, regardless of formal hierarchy. The senior officer defers to the technical expert when the technical question dominates.

Resilience engineering builds on HRO with explicit attention to the dynamics of safety: the system's ability to adjust its functioning prior to, during, and following changes and disturbances. The Resilience Analysis Grid (Hollnagel) characterises four cornerstones: respond (to disturbances when they occur), monitor (for what is happening), learn (from experience), anticipate (potential future disturbances).

The engineering application: design safety management systems that institutionalise the HRO characteristics; design operational practices that embody resilience; design organisational learning processes that turn experience into improvement. The investment is substantial; the payoff is operations that are safer than the inherent risk would predict.

\section{Failure as data, not as shame}\pathtag{core}

The cultural disposition toward failure is decisive. Organisations that treat failure as a source of data investigate it openly, share what is learned, and improve. Organisations that treat failure as shame to be hidden suffer worse failures because the early warnings are suppressed.

The blameless post-mortem is the institutional expression of this disposition: the post-mortem focuses on what happened, why, and how to prevent recurrence; it does not focus on who to blame. The discipline is not "no consequences" (gross negligence and malice still warrant accountability); it is "blame is the wrong lens" for the engineering improvement that the post-mortem should produce.

The discipline of treating failure as data extends beyond the post-mortem:

\textbf{Open reporting:} workers can report near-misses, defects, and concerns without fear. The Federal Aviation Administration's Aviation Safety Reporting System (NASA-administered) is a model: voluntary, confidential, with limited immunity. The data informs safety regulation and operational improvement.

\textbf{Cross-organisational sharing:} learnings from one organisation's failures benefit others. Aviation, nuclear, and medical-device industries have formal mechanisms for cross-organisation sharing; many other industries do not.

\textbf{Sustained investigation:} the investigation is funded and staffed to complete; not truncated when politically convenient or when blame is assigned. The NTSB (National Transportation Safety Board, USA) is a model: independent, well-funded, methodologically rigorous.

\textbf{Long-term tracking:} repeat occurrences of similar failures indicate that learning has not happened. The discipline includes tracking failures across years, looking for patterns.

\textbf{Cultural support:} the organisation's leadership signals that failure data is welcome. The signal is communicated through resource allocation, acknowledgement of failure-related learning, and protection from retaliation for raising concerns.

The opposite disposition (treating failure as shame) is common and produces measurable harm. Schedules push out near-misses; the failure rate rises until the catastrophic event occurs; the investigation finds many missed warnings; the cycle repeats. The 2003 Columbia disaster (\cite{acc:caib-columbia}) had a strong "failure-as-shame" component: foam-shedding events had been classified as "in family" rather than as a developing problem; the engineering concerns about wing damage on the Columbia flight were not effectively communicated to management.

\section{The reader's posture: how to study failure without becoming useless}\pathtag{standard}

Studying failure can produce two unhelpful outcomes: paralysis (the conviction that nothing can be safely built) or cynicism (the conviction that all systems are doomed and effort is futile). The productive posture is between these: rigorous study of failure mechanisms combined with disciplined optimism about engineering capability.

The constructive disposition includes:

\textbf{Acknowledge the failure rate.} All systems fail at some rate. The engineering question is the rate, the conditions, the consequences, the mitigation. The right rate is not zero; the right rate balances cost of prevention against cost of failure.

\textbf{Focus on learning.} Each failure is an input. The engineer's job is to extract the engineering lesson, generalise it across contexts, and apply it forward.

\textbf{Avoid hindsight bias.} What looks obvious after the fact was not obvious before. The investigation should evaluate decisions in the context the decision-makers actually had, not in light of subsequent knowledge.

\textbf{Avoid blame-the-victim reasoning.} The operator's error often had organisational, design, and procedural contributors; the design that was easy to use wrong often was, in fact, used wrong. The engineer's analysis identifies these contributors rather than stopping at "user error."

\textbf{Build for failure tolerance.} Since failure is inevitable, engineer the system to fail safely: graceful degradation, redundancy, recovery procedures, learning capability. Failure tolerance is a design property; it is achieved through engineering effort.

\textbf{Maintain perspective on success.} Most operations succeed. The engineering disciplines that produce success are also part of the picture; failure study should not crowd them out.

The reader who maintains the constructive disposition can read failure case studies productively, derive lessons that generalise, and apply them forward without becoming pessimistic about the discipline or oneself.

\section{Failure: failures of failure analysis}\pathtag{core}

Failure analysis itself can fail; the discipline has its own failure modes.

\textbf{Stopping at superficial cause:} the analysis identifies a proximate cause and stops there, missing the deeper contributors. "Operator error" is the canonical stopping point; the design, procedure, training, and organisation that produced the error are not investigated. Mitigation: the "five whys" technique; structured investigation methods (TRIPOD, MTO, AcciMap, STAMP).

\textbf{Hindsight bias:} the analysis evaluates the decisions in light of the outcome the analysts now know. The decision-makers' actual information and constraints are not appreciated. Mitigation: explicit reconstruction of the decision context; "fundamental attribution error" awareness.

\textbf{Blame allocation:} the analysis serves to allocate blame rather than to identify systemic improvements. The analysis is shaped by legal and political considerations; the engineering improvement is secondary. Mitigation: institutional separation between investigation and adjudication; safety-management systems with explicit blameless mandates.

\textbf{Single-cause framing:} the analysis fixes on a single cause; the contributing factors are minimised or hidden. Mitigation: explicit acknowledgement of multiple causes; structured analysis methods.

\textbf{Lessons not implemented:} the analysis identifies the right lessons; the organisation does not implement them. The shelf of post-mortem reports without implementation is the cultural failure mode. Mitigation: closure of recommendations; tracking of implementation; periodic audit.

\textbf{Lessons not generalised:} the analysis treats the failure as unique to its context; the cross-context applicability is missed. The same failure mechanism then appears in different contexts. Mitigation: cross-domain literature; structured taxonomy of failure mechanisms; communities of practice across organisations.

\textbf{Lessons lost over time:} the lessons are documented but the organisational memory fades; new staff are unaware. The same failure recurs decades later. Mitigation: institutional memory practices; periodic refresher training; ongoing case-study libraries.

\textbf{Lessons distorted in transmission:} the lesson as remembered differs from the lesson that should have been learned. Cultural pressures shape the memory; convenient interpretations crowd out inconvenient ones. Mitigation: original documents preserved; independent review of the lessons over time.

\textbf{Investigation captured by stakeholders:} the investigation is biased by interested parties (the regulator, the operator, the manufacturer, the insurer). The findings serve the stakeholder rather than the engineering. Mitigation: independence of investigation bodies; multiple-party investigation; transparency.

The discipline of investigation is itself a discipline; the engineer treats failure analysis with the same care as failure design.

\begin{principle}[The failure-as-engineering-subject principle]
Failure is an engineering subject with its own mechanisms, statistics, methods, and discipline. The engineer studies failure systematically: characterise the mechanism, quantify the rate, identify the precursors, design the mitigation, monitor for the precursors, respond to events, learn from each, share with the community. The half-discipline (study only success) is incomplete; the full discipline (study both) is engineering. The investment in failure study is not overhead; it is the work that makes engineering reliable, safe, and improving.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Read three primary sources on a historical accident]
\textbf{Objective.} Pick a historical engineering accident; read three primary sources about it; identify what each missed; develop the engineer's own assessment of the lessons.

\textbf{Method.} Choose an accident with substantial documentation: Challenger (1986), Three Mile Island (1979), Bhopal (1984), Chernobyl (1986), Tacoma Narrows (1940), Hyatt Regency (1981), Therac-25 (1985-1987), Mars Climate Orbiter (1999), Columbia (2003), Air France 447 (2009), Fukushima (2011), Boeing 737 MAX (2018-2019), or another for which primary investigation reports are publicly available. Read at least three primary sources: the official accident report, a technical engineering analysis, and a popular-press treatment (or alternative formats). For each source, identify what it captures well and what it misses or distorts. Synthesise the engineering lessons in your own words.

\textbf{Deliverables.} The accident description; the three sources with full citation; analysis of each source (strengths, gaps, distortions); your synthesis of the engineering lessons.

\textbf{Hazard.} Reading accident reports can be emotionally affecting, particularly for accidents involving loss of life. The educational value depends on engaging seriously with the material rather than dismissing it.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A safety-management system tracks $20$ defensive layers, each with $5\%$ failure probability per opportunity. Compute the probability that all layers fail simultaneously, assuming independence.
\end{exercise}

\begin{exercise}
Two failure modes have probabilities $p_1 = 10^{-4}$ and $p_2 = 5 \times 10^{-5}$ per operation; they are independent. Compute the probability of either or both occurring in $10^4$ operations.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the relationship between near-miss frequency and accident frequency under the Heinrich-pyramid hypothesis.
\end{exercise}

\begin{exercise}
Derive the conditions under which the Swiss-cheese model's independence assumption fails (common-mode failures across layers).
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the cost of an inadequate post-mortem (lessons not learned) for a moderately-large engineering organisation over a decade.
\end{exercise}

\begin{exercise}
Estimate the staffing and resourcing required for a robust safety-management system in a high-reliability industrial operation.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate the Heinrich-pyramid statistics: many near-misses, fewer minor incidents, rarely accidents. Compare to empirical data from a chosen industry.
\end{exercise}

\begin{exercise}
Simulate the effect of independence-violation on accident rates: how does common-mode failure across nominally independent defensive layers change the accident probability.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a safety-management framework for a startup developing an autonomous mobile robot for hospital use. State the key processes, the staffing, the regulatory interface.
\end{exercise}

\begin{exercise}
Design a near-miss reporting system for an industrial operation: anonymity, reporting channels, classification, investigation, follow-up.
\end{exercise}

\begin{exercise}
Design the curriculum for a senior engineer's "failure literacy" training: case studies, methods, communication skills, organisational dynamics.
\end{exercise}

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A team's post-mortem identifies operator error as the root cause; the operator has since left the company. State the engineering response and the role of system-level analysis.
\end{exercise}

\begin{exercise}
A team's near-miss reporting rate has dropped significantly; the team's leadership interprets this as improvement. State the engineering interpretation and the alternative explanation of reporting suppression.
\end{exercise}

\begin{exercise}
A team is asked to perform a post-mortem on a recent failure; the team's management has already publicly assigned blame. State the engineering response and the role of independent investigation.
\end{exercise}

\begin{exercise}
A team is choosing between a Swiss-cheese-model framework and a STAMP-based investigation method for a recent incident. State the engineering criteria.
\end{exercise}

\begin{exercise}
A team's safety record has improved every year for a decade; the team's leadership believes the safety system is mature. State the engineering response and the role of preoccupation with failure.
\end{exercise}

\begin{exercise}
A team's post-mortem recommendations have been implemented; a recurrence of a similar failure has occurred. State the engineering response.
\end{exercise}

\begin{exercise}
A team's organisation is under schedule pressure; safety-related concerns have been deferred. State the engineering response and the role of HRO principles.
\end{exercise}

\begin{exercise}
A team is asked to write a public-facing description of a recent failure for an internal newsletter. State the engineering posture, the role of accountability without blame, and the educational value of the content.
\end{exercise}

\begin{exercise}
A team is participating in a cross-organisation failure-data sharing program; the team's organisation has reservations about disclosing certain incidents. State the engineering analysis and the trade-offs.
\end{exercise}
