% Volume 2: Form
% Opener: tagline + scope. Sits immediately after the \part{} declaration.

\vspace*{1.5cm}
\begin{center}
\itshape\large
Mathematics as the language of engineering
\end{center}
\vspace{0.5cm}

\noindent
We teach mathematics for engineers: usable power, not disciplinary completeness. The reader leaves able to differentiate, integrate, solve linear systems, set up and solve ordinary and the simpler partial differential equations, reason probabilistically, optimise under constraint, and translate a real problem into a model.

\vspace{0.5cm}

\section*{What this volume is for}

Volume I taught the reader that engineering begins with measurement. Volume II teaches the reader the language in which the rest of the work is written. Mathematics, here, is not a parallel discipline that engineers happen to use. It is the engineer's working notation. A reader who arrives at Volume III without a fluent grasp of derivatives, integrals, vectors, matrices, eigenvalues, ODEs, the simpler PDEs, probability, and optimisation will read Volumes III through XII as a sequence of unrelated theorems. A reader who spends time here will read the rest as a sequence of decisions about how a quantity moves, accumulates, balances, scales, breaks, and recovers.

The volume is the largest of the twelve, by chapter count and by page count, and the standard is engineering mathematics. By that we mean: usable power, fluency under time pressure, a sense of which approximation costs the engineer how much, and the discipline to refuse a method when its preconditions are not met. The reader does not leave able to prove the spectral theorem from measure-theoretic foundations. The reader leaves able to use the singular-value decomposition on a noisy data matrix, identify when its conditioning is bad, and choose a numerical strategy that does not amplify the noise.

\section*{What the reader brings}

The reader who opens Volume II has finished Volume I and carries its habits: estimation before calculation, units through every step, error as a first-class quantity, calibration discipline, statistical literacy at the engineering level. Mathematically, the reader brings algebra, geometry, trigonometry, and the rudiments of probability and statistics introduced in Volume I Chapter 8. The reader does not bring calculus. Calculus arrives in this volume.

The reader is also assumed to bring physical intuition for the quantities the mathematics is about. A derivative will be introduced as a rate, but the reader is expected to know what a rate is from Volume I and from ordinary life. An integral will be introduced as accumulation, but the reader is expected to know what accumulation is. A vector will be introduced as a directed quantity, but the reader will already have met velocity, force, and current as physical examples. The mathematics formalises intuitions the reader already has; it does not invent them.

A reader who attempts this volume without the Volume I habits will find that the mathematics works on the page but does not connect to anything the reader can measure or test. We have written each chapter with the assumption that the reader will, periodically, stop and run a small numerical check, sketch a quick estimate, or verify a unit balance. Those habits are what turn the formal apparatus into engineering mathematics.

\section*{The arc of the eighteen chapters}

The volume is organised in five arcs.

\textbf{Arc one: foundations} (Chapters 1-4). Algebraic identities, inequalities, proof discipline, and notation as the syntax. Functions as mappings, polynomials, exponentials, logarithms, and trigonometry as the working vocabulary. Complex numbers and phasors as the engineer's rotation tool. Sequences, series, limits, and Taylor expansion as the bridge into calculus. The arc closes with the small-angle approximation, the geometric series, and the engineer's first formal estimate of a function near a working point.

\textbf{Arc two: calculus} (Chapters 5-7). Differentiation as the first formal handling of rates. Integration as the formal handling of accumulation. Ordinary differential equations as the formal handling of dynamics. Each chapter takes the apparatus through to the engineer's working application: optimisation under one variable, work and moment of inertia, the RC circuit and the pendulum. Numerical methods enter alongside the analytical methods, in keeping with the discipline of using mathematics on real data.

\textbf{Arc three: linearity in many variables} (Chapters 8-11). Vectors as algebraic and geometric objects. Linear algebra as the language of systems with many degrees of freedom. Eigenvalues and decompositions as the structural understanding of linear maps. Multivariable calculus as the language of fields, gradients, and surfaces. The arc culminates in the divergence and curl theorems, which connect the local and the global through the same balance discipline introduced in Volume I.

\textbf{Arc four: distributed and uncertain} (Chapters 12-14). Partial differential equations as the language of distributed phenomena (heat, waves, Laplace fields). Probability theory as the language of uncertainty deeper than measurement error. Statistical inference as the engineer's tool for deciding what a finite sample says about an underlying population, building on the engineering-statistics chapter of Volume I.

\textbf{Arc five: applied mathematics} (Chapters 15-18). Optimisation as the working discipline of choosing among feasible alternatives. Numerical methods as the discipline of running the previous chapters on a finite-precision machine. Discrete mathematics as the literacy required to discuss algorithms, graphs, and combinatorics with later-volume material on information, machines, and systems. The volume closes with mathematical modelling: a working chapter on how to take a real engineering question and turn it into the apparatus this volume has spent fifteen-hundred pages building.

\section*{What the reader will be able to do at the end}

A reader who finishes Volume II should be able to:

\begin{itemize}[leftmargin=*]
  \item Differentiate and integrate elementary functions, including
    standard families and worked-rule combinations, fluently and
    without lookup, and identify when an integral does not have a
    closed form.
  \item Set up and solve a first-order or second-order linear ODE,
    including the forced-response case, and identify when a numerical
    solver is required.
  \item Solve a linear system by Gaussian elimination, LU, or QR,
    and identify when a system is ill-conditioned and what to do
    about it.
  \item Compute eigenvalues and eigenvectors of small matrices and
    interpret the singular-value decomposition of a data matrix.
  \item Compute gradients, divergences, and curls in standard
    coordinate systems and apply the integral theorems.
  \item Recognise the heat equation, the wave equation, and the
    Laplace equation as the canonical PDEs of physical engineering.
  \item Compute probabilities and conditional probabilities, identify
    standard distributions, and design a small statistical inference
    procedure.
  \item Set up and solve a small constrained optimisation problem
    and identify when the problem is convex.
  \item Implement and analyse a numerical method (root-finding,
    quadrature, ODE integration, linear-system solver) with proper
    accounting of truncation and floating-point error.
  \item Take a real engineering question, identify which of the
    above tools applies, and produce a model that an outside
    engineer could review.
\end{itemize}

A reader who reaches Volume III without these capabilities has read Volume II without doing it. We say so plainly because Volume III's first page assumes the language is already fluent.

\section*{Archetypes introduced or developed in this volume}

Volume I introduced six archetypes (scaling, balance, failure, uncertainty, interface, transport). Volume II introduces three more and develops the existing six.

\begin{itemize}[leftmargin=*]
  \item \engterm{Optimisation}: introduced in Chapter 5 as the
    one-variable case (find a maximum or minimum), developed in
    Chapter 11 as the multivariable case (gradient equals zero,
    Lagrange multipliers), and given its formal home in Chapter 15.
    Optimisation recurs in every later volume that asks the
    engineer to choose among feasible alternatives.
  \item \engterm{Stability}: introduced in Chapter 3 as rotation
    preserving structure, developed in Chapter 7 as the qualitative
    behaviour of ODE solutions, and revisited in Chapter 10 as the
    eigenvalue condition for linear-system stability. The archetype
    recurs across Volumes III, IV, IX, and X.
  \item \engterm{Control}: introduced informally in Chapter 7 as
    the forced response of a second-order system, ready for full
    development in Volumes VIII and IX.
\end{itemize}

The other archetypes (scaling, balance, failure, uncertainty, interface, transport) are not introduced fresh, but each is given mathematical apparatus here that it lacked in Volume I. Scaling becomes Taylor expansion and asymptotic series. Balance becomes the divergence theorem. Failure becomes numerical instability and ill-conditioning. Uncertainty becomes probability theory. Interface becomes boundary conditions. Transport becomes the gradient and the convection-diffusion equation.

\section*{The volume project}

The volume project is described in detail in Chapter 18. The short form: the reader takes a real-world phenomenon of their choice (a heating system, a queue, a small electrical circuit, a population of bacteria, anything they can measure with the Volume I instruments) and produces a 20-30 page document containing data, model derivation, model fit, error analysis, predictions, and validation. The document is judged against the same rubric a mid-career engineer would face: would the model survive a hostile review?

The project is the volume's deliverable, and the credential by which the reader earns admission to Volume III, where the language of this volume is taken for granted.

\section*{A note on the next volume}

Volume III takes up Force: statics, dynamics, vibrations, fracture, and the working physics of the structures and systems an engineer designs. The bridge is direct. Every chapter of Volume III opens with a vector, a derivative, an ODE, or a linear system. The reader who carries the habits of this volume across that bridge has the language of physical engineering. The reader who arrives at Volume III with the apparatus but without the habits will read it as theorem followed by theorem; the reader who arrives with both will read it as a sequence of decisions about real, dated, error-bearing forces acting on real, dated, error-bearing bodies.

\vspace{1cm}
