% Volume 10: Failure
% Opener: tagline + scope. Sits immediately after the \part{} declaration.

\vspace*{1.5cm}
\begin{center}
\itshape\large
A real engineer is someone who knows how things die
\end{center}
\vspace{0.5cm}

\noindent
The first nine volumes built engineering as the discipline of making reliable intervention under constraint. Volume X faces the discipline's adversary directly: failure. Every chapter in the previous volumes carried a failure section because failure is unavoidable; this volume gives failure its full treatment as a first-class engineering subject, with the same rigour given to design, control, and operation in the volumes that preceded it. The thesis is plain: an engineer who studies only success understands half the work; the other half is the systematic study of how things break.

The volume's scope is broad. Mechanical failure (fatigue, fracture, corrosion, wear) and dynamical failure (instabilities, runaway, buckling, resonance) cover the physical mechanisms. Software defects, concurrency hazards, and cascading network failure cover the discrete and distributed mechanisms. Common-mode failure, human factors, organisational pathology, and regulatory failure cover the cross-cutting modes that defeat the redundancy and diversity that classical reliability theory predicts. Forensic engineering, the discipline of failure investigation, gives the methods. Near-miss analysis catches what would have failed if conditions had aligned differently. Maintenance failure addresses the slow accident that grows from deferred attention. Climate, ecological, and planetary failure scales the discipline to consequences that match the engineering's reach. The closing chapter develops the discipline of post-mortem: the systematic, blameless review that turns failure into the organisation's most valuable input.

The arc moves from physical and software mechanisms (Ch 2-6) through systemic and human failure modes (Ch 7-11) through investigation, near-misses, and maintenance (Ch 12-14) to large-scale and procedural concerns (Ch 15-16). The reader who finishes this volume has a working vocabulary for analysing how engineered systems fail, the discipline for investigating failures honestly, the patterns to recognise across domains, and the perspective to design for failure tolerance rather than for the illusion of failure prevention.

The archetypes that emerged in earlier volumes intensify here: failure (every chapter); coupling (failures cross subsystem boundaries); uncertainty (the conditions under which failure occurs are often unknown until they happen); balance (the cost of preventing failure vs the cost of failing); scaling (small failures and large failures follow different patterns). The new archetype this volume introduces is investigation: the disciplined retrospective that separates what happened from what was supposed to happen.

\TODO{Volume 10 opener: ~5pp in finished form. The above lands the volume's thesis, scope, arc, and archetype family; the production-density pass adds the bridge from Vol IX, the explicit promise list, and an opening case study that crosses several of the failure modes (Bhopal, Fukushima, or the Challenger disaster as the multi-aspect example that prefigures the chapter sequence).}

\vspace{1cm}
