\chapter{Bioreactors and biomanufacturing}
\label{vol06:ch10}

\epigraph{The product of a biological manufacturing process is the
state of a population of living cells at a particular time, captured
in a particular container, with a particular record.}{paraphrase of
\cite{text:v6c10-shuler-bioprocess-2017}, ch.~1}

\chapmeta{Half-life: medium-life. Archetypes: balance, transport,
scaling. Exercise target: 25. Project track: household fermentation
plus analysis (hybrid). Hazard class: low (food-handling hygiene,
glassware, hot water; no pathogen handling and no pressure vessels).
Reader path default: standard, with sections explicitly tagged. Named
cases: Genzyme Cerezyme Vesivirus 2117 contamination, 2009 (Allston
Landing, Massachusetts).}

\noindent
Biomanufacturing is the engineering of a living production system. A
chemical plant asks molecules to react; a bioprocess asks cells to grow,
maintain themselves, express the product, survive the stress imposed by
the product, and remain genetically stable long enough for the run to be
worth keeping. The reader who finishes this chapter can read a
bioreactor as a balance around biomass, substrate, oxygen, heat, and
product; can name the scale-up variables that do not travel cleanly from
flask to factory; can distinguish microbial, mammalian, and plant
production systems; can trace one industrial example, recombinant human
insulin, from production strain through fed-batch fermentation, cell
lysis, refolding, ion-exchange and reversed-phase polishing,
crystallisation, formulation, and fill; and can explain why the
downstream purification train often costs more attention than the
reactor itself.

\section{Bioreactor types}\pathtag{core}

A \engterm{bioreactor} is a bounded vessel or device in which living
cells, enzymes, or tissues perform a controlled biochemical
transformation. The boundary matters. Feed enters; product, waste, heat,
and gas leave; cells remain, leave, or are recycled depending on the
process. The engineer's first task is to draw that boundary and state
which quantities cross it.

\input{volumes/06-life/ch10-bioreactors-and-biomanufacturing/figures/fig-bioreactor-types.tex}

\engterm{Batch reactors}. In a batch culture, the vessel is charged
with medium and inoculum, then run until a harvest condition is reached.
No fresh medium enters during the main production phase. The batch is
simple, inspectable, and easy to isolate from contamination. It is also
finite: the substrate is consumed, inhibitory products accumulate, and
the cells pass through lag, exponential, stationary, and decline phases.
Batch operation fits products tied to a finite growth cycle, processes
where contamination risk dominates, and products whose purification is
organised around one harvest at a time
\cite[ch.~3]{text:v6c10-doran-bioprocess-2013}.

\engterm{Fed-batch reactors}. Fed-batch operation adds feed during the
run while withholding continuous harvest. The feed rate becomes a
control variable. It can limit substrate to avoid overflow metabolism,
extend the production phase, hold the cells away from inhibitory
concentrations, and tune the growth rate. Fed-batch is the default
industrial compromise for many microbial and mammalian processes:
more controllable than batch, less operationally exposed than continuous
culture \cite[ch.~12]{text:stanbury-fermentation}.

\engterm{Continuous reactors}. In a continuous stirred-tank bioreactor,
fresh medium enters at volumetric flow rate $F$ and culture leaves at
the same rate. The working volume $V$ is constant, and the
\engterm{dilution rate} is
\[
D = \frac{F}{V}.
\]
At steady state, the cells grow at the dilution rate. A chemostat uses
this fact to make growth rate a setpoint. It is a beautiful machine for
studying physiology and evolution; it is a demanding production machine
because contamination, mutation, and drift have indefinite time to act.

\engterm{Perfusion reactors}. In perfusion culture, fresh medium is
fed continuously and product-containing spent medium is removed while
cells are retained by a filter, settler, centrifuge, or membrane. The
cell density can rise far above ordinary batch culture. Perfusion fits
fragile high-value products, including some mammalian-cell products,
where keeping cells in a productive state beats repeated batch start-up.
The price is a more complicated boundary: medium, product, cells, and
waste no longer share the same residence time.

\engterm{Stirred tanks, air-lift reactors, packed beds, and
photobioreactors}. The tank geometry is not decorative. A stirred tank
uses impellers and baffles to mix liquid and disperse gas. An air-lift
reactor uses gas injection and density differences to circulate the
broth with lower shear. A packed bed immobilises cells or enzymes on a
solid support and passes medium through the voids. A photobioreactor
adds light as an input and makes surface area, shading, and heat removal
central design constraints. The vessel type is chosen from the transport
problem, not from a catalogue preference
\cite[ch.~8]{text:v6c10-shuler-bioprocess-2017}.

\engterm{Single-use systems}. Many modern bioprocesses use disposable
bags, tubing assemblies, filters, and sensors for development and some
production scales. Single-use systems reduce cleaning validation burden
and speed changeover, but they introduce extractables, leachables,
plastic-film oxygen transfer limits, waste handling, and supplier
dependence. They are engineering tradeoffs, not a general upgrade. The
current market data show single-use stirred-tank reactors well
established at the 200 L to 2000 L production scale for clinical and
some commercial mammalian-cell processes
\cite{paper:v6c10-walsh-biopharmaceuticals-2022}.

\engterm{Choosing among the types}. The choice of reactor type is a
function of the cell platform, the product, the regulatory category,
the production volume, and the existing facility. A short tabular
heuristic, current as of mid-2020s industrial practice:

\begin{itemize}[leftmargin=*]
\item Fast microbial heterotroph, secreted enzyme, large volumes,
  commodity prices: stirred-tank, fed-batch, hard-piped, steam-in-place.
\item Filamentous fungus, secondary metabolite, high oxygen demand,
  shear-sensitive morphology: stirred-tank with low-shear impeller, or
  air-lift, fed-batch.
\item CHO monoclonal antibody, clinical to commercial scale: single-use
  or hard-piped stirred-tank fed-batch (200 L to 20000 L); perfusion
  considered when titer plateau persists or when intensification gives
  better facility utilisation.
\item Mammalian-cell viral vector, gene therapy, small batch:
  single-use stirred-tank, often perfusion or intensified fed-batch.
\item Plant or algal cell culture for natural products: photobioreactor
  or large open pond, light limited.
\item Whole-cell biocatalysis, immobilised enzyme: packed bed,
  fluidised bed, or continuous stirred tank, depending on the
  hydrodynamics and the cell or enzyme tolerance to shear.
\end{itemize}

The heuristic is not a substitute for the engineering calculation. It
is a list of starting points around which the calculation can run.

\begin{archetype}[Balance: the living phase is inside the account]
The biological phase is not a catalyst drawn as a symbol above the
reaction arrow. It is part of the mass, energy, information, and risk
account. Biomass consumes carbon, nitrogen, oxygen, phosphate, trace
metals, and operator attention. It produces heat, carbon dioxide, acid
or base load, secreted product, intracellular product, extracellular
polymer, and variants of itself. A reactor balance that leaves the
living phase implicit has already lost the process.
\end{archetype}

\section{Mass and heat balances in bioreactors}\pathtag{core}

The generic bioreactor balance is the same balance used throughout this
book:
\[
\frac{d(\text{amount in vessel})}{dt}
= \text{in} - \text{out} + \text{generation} - \text{consumption}.
\]
The difference is that the generation and consumption terms are produced
by a living population whose state changes during the run.

\subsection{Growth kinetics}

Let $X$ be biomass concentration, $S$ substrate concentration, and $P$
product concentration. A common first model for microbial growth uses
Monod kinetics,
\[
\mu(S) = \mu_{\max}\frac{S}{K_S + S},
\]
where $\mu$ is the specific growth rate, $\mu_{\max}$ is the maximum
specific growth rate, and $K_S$ is the substrate concentration at which
$\mu = \mu_{\max}/2$ \cite{paper:v6c10-monod-1949}. The form is the
same saturating shape as Michaelis-Menten enzyme kinetics
\cite{paper:v6c10-michaelis-menten-1913}. The interpretation differs: a
whole cell grows, regulates, and reallocates resources, so $K_S$ is an
empirical parameter, not a single enzyme constant.

\input{volumes/06-life/ch10-bioreactors-and-biomanufacturing/figures/fig-monod-growth.tex}

For a well-mixed batch culture with one limiting substrate,
\begin{align*}
\frac{dX}{dt} &= \mu(S)X, \\
\frac{dS}{dt} &= -\frac{1}{Y_{X/S}}\mu(S)X - m_S X, \\
\frac{dP}{dt} &= q_P X.
\end{align*}
Here $Y_{X/S}$ is the biomass yield on substrate, $m_S$ is a
maintenance substrate demand, and $q_P$ is a specific product-formation
rate. The equations are already an engineering judgement. They assume
one limiting substrate, no inhibition, homogeneous mixing, no cell
death, and a product rate independent of growth phase. Those assumptions
are acceptable for a first calculation and dangerous as a release
specification.

\subsection{Monod kinetics worked}

Take \emph{E.\ coli} on glucose with $\mu_{\max} = 0.90\,\text{h}^{-1}$
and $K_S = 0.050\,\text{g}/\text{L}$. The specific growth rate is
within 5 percent of $\mu_{\max}$ as long as $S$ is more than about
$1\,\text{g}/\text{L}$, since
\[
\frac{\mu}{\mu_{\max}}
= \frac{S}{K_S + S}
= \frac{1.0}{0.050 + 1.0}
= 0.952.
\]
A batch starting at $S_0 = 20\,\text{g}/\text{L}$ therefore spends
most of its course in the near-saturating regime. The substrate falls
to $K_S$ only near the end of the exponential phase, after which the
growth rate falls sharply and the culture enters a substrate-limited
deceleration. The same expression captures the physiological case
that matters for fed-batch design: by holding $S$ near $0.2\,\text{g}/
\text{L}$ a feed controller picks a specific growth rate of
$\mu = 0.90 \times 0.2 / (0.05 + 0.2) = 0.72\,\text{h}^{-1}$, that is,
80 percent of $\mu_{\max}$ at one-hundredth of the batch-starting
substrate. The reduction reduces overflow metabolism (acetate in
\emph{E.\ coli}, ethanol in \emph{S.\ cerevisiae} under the Crabtree
effect) and is the central motivation for fed-batch industrial
fermentation.

The fit of $\mu_{\max}$ and $K_S$ to laboratory data is itself an
engineering operation. The `monod_growth_fit.py` script in `code/`
performs a non-linear least-squares fit to synthetic data and reports
parameters with confidence intervals; the same form runs on real
optical-density or carbon-dioxide-evolution data after conversion to
specific rates.

\subsection{Biomass yield $Y_{X/S}$ derivation}

The yield coefficient $Y_{X/S}$ is the ratio of biomass produced to
substrate consumed under reference conditions. For aerobic
heterotrophic growth on glucose, a simple stoichiometric construction
follows from a generalised cell composition $\text{CH}_{1.8}\text{O}_{0.5}\text{N}_{0.2}$,
which has a molar mass of about $24.6\,\text{g}/\text{C-mol}$ and a
degree of reduction $\gamma_X \approx 4.2$. Glucose
$\text{C}_6\text{H}_{12}\text{O}_6$ has a molar mass
$180\,\text{g}/\text{mol}$, a C content of $6\,\text{C-mol}/\text{mol}$,
and a degree of reduction $\gamma_S = 4$.

A mass balance with no product formation and an aerobic generalised
stoichiometry,
\[
\text{C}_6\text{H}_{12}\text{O}_6 + a\,\text{O}_2 + b\,\text{NH}_3
\longrightarrow c\,\text{CH}_{1.8}\text{O}_{0.5}\text{N}_{0.2}
+ d\,\text{CO}_2 + e\,\text{H}_2\text{O},
\]
closes carbon, hydrogen, oxygen, nitrogen, and electron balance with
five equations in the five stoichiometric coefficients $a, b, c, d, e$
once a sixth constraint (the experimentally measured respiratory
quotient, oxygen uptake, or biomass yield) is supplied. The
electron-balance equation,
\[
6\gamma_S + b \cdot 0 - 4a = c\,\gamma_X + d \cdot 0,
\]
with $\gamma_S = 4$ and $\gamma_X = 4.2$, gives an algebraic
relationship between biomass and the oxygen demand. The result, for a
glucose-limited aerobic culture with $a = 1.5$ mol $\text{O}_2$ per mol
glucose (a typical mid-range value), is $c \approx 2.8$ C-mol biomass
per mol glucose. Converting to mass:
\[
Y_{X/S}
= \frac{2.8 \times 24.6\,\text{g/C-mol}}{180\,\text{g/mol}}
= 0.38\,\text{g/g}.
\]
The order-of-magnitude $0.4\,\text{g/g}$ that pervades the literature
is a derived number, not a fitted one. Higher oxygen uptake ($a$ larger)
pushes more carbon to CO$_2$ and reduces $c$; lower oxygen uptake leaves
more carbon in biomass but cannot be sustained because the electron
balance forces overflow metabolism (acetate, lactate, ethanol) and the
culture loses growth efficiency. The yield-coefficient table at
`data/yield_coefficients.csv` collects representative $Y_{X/S}$ and
$Y_{P/S}$ values for the platform comparisons in \S\ref{sec:vol06ch10:platforms}.

\subsection{Continuous culture and washout}

For a chemostat with feed substrate concentration $S_f$ and dilution
rate $D$, the balances become
\begin{align*}
\frac{dX}{dt} &= \left(\mu(S) - D\right)X, \\
\frac{dS}{dt} &= D(S_f - S) - \frac{1}{Y_{X/S}}\mu(S)X.
\end{align*}
At nonzero steady state, $\mu(S^\ast)=D$. The reactor turns growth
rate into a knob. Increase $D$, and the culture must grow faster to
remain in the vessel. If $D$ exceeds the maximum growth rate available
at the feed and operating conditions, biomass washes out and the reactor
becomes a sterile pipe.

The washout condition is not a mathematical curiosity. It is the
operating limit of continuous culture. A shift in pH, temperature,
inhibitor concentration, phage burden, or trace nutrient can lower
$\mu_{\max}$ enough that a dilution rate that was safe yesterday
becomes washout today. The operator sees the failure first as falling
optical density, falling oxygen uptake, and product loss.

Solving the steady-state Monod chemostat algebra gives
\[
S^\ast = \frac{K_S D}{\mu_{\max} - D},
\quad
X^\ast = Y_{X/S}(S_f - S^\ast),
\]
valid only while $D < \mu_{\max}$. At $D = \mu_{\max}$, $S^\ast = S_f$
and $X^\ast = 0$: total washout. Maximum biomass productivity
$D \cdot X^\ast$ occurs at an intermediate dilution rate
\[
D_{\text{opt}} = \mu_{\max}\left(1 - \sqrt{\frac{K_S}{K_S + S_f}}\right),
\]
which for $S_f \gg K_S$ is close to but not at $\mu_{\max}$. The
operating point for an industrial continuous fermenter therefore
balances productivity against the failure margin to washout.

\subsection{Yield accounting}

A yield is a ratio that only makes sense with its basis named. Biomass
yield on substrate, $Y_{X/S}$, has units such as
$\text{g biomass}/\text{g glucose}$. Product yield on substrate,
$Y_{P/S}$, has units such as $\text{g product}/\text{g glucose}$.
Productivity is a rate, often $\text{g}/(\text{L}\cdot\text{h})$.
Titer is a concentration, often $\text{g}/\text{L}$. A high titer can
come with poor productivity if the run takes too long. A high
productivity can come with poor final titer if the product is harvested
too dilute. A high yield can still be uneconomic if downstream recovery
destroys the product.

The reactor calculation therefore carries at least four quantities:
titer, rate, yield, and recovery. A process that reports only one of
them is hiding the tradeoff.

\subsection{Heat balance}

Cells dissipate heat. Aerobic growth turns reduced carbon into biomass,
CO$_2$, water, product, and heat. Anaerobic fermentation releases less
heat per mole of substrate because more chemical free energy remains in
the product. In either case, a reactor that cannot remove heat cannot
hold its biological setpoint.

\input{volumes/06-life/ch10-bioreactors-and-biomanufacturing/figures/fig-aerobic-anaerobic.tex}

A simple heat balance for a well-mixed jacketed reactor is
\[
\rho C_p V \frac{dT}{dt}
= \dot{Q}_{\text{met}} + \dot{Q}_{\text{agitation}}
- U A (T - T_c) - F\rho C_p (T - T_f),
\]
where $T$ is broth temperature, $T_c$ jacket coolant temperature, $T_f$
feed temperature, $U A$ the heat-transfer conductance, and
$\dot{Q}_{\text{met}}$ the metabolic heat release. The agitation term
matters at large scale: impeller power eventually becomes heat in the
broth.

\begin{estimation}
\textbf{A $1000\,\text{L}$ aerobic microbial reactor consumes oxygen
at $0.10\,\text{mol}/(\text{L}\cdot\text{h})$. Estimate the metabolic
heat that must be removed.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The oxygen uptake is $100\,\text{mol}/\text{h}$ for the whole vessel.
One mole of O$_2$ is $32\,\text{g}$, so this is
$3.2\,\text{kg}/\text{h}$ of oxygen. Oxidising reduced carbon releases
about $14\,\text{MJ}$ per kilogram of O$_2$ consumed, at order of
magnitude. The heat load is therefore
\[
3.2\,\frac{\text{kg O}_2}{\text{h}}
\times 14\,\frac{\text{MJ}}{\text{kg O}_2}
\approx 45\,\text{MJ}/\text{h}
\approx 12.5\,\text{kW}.
\]
A household kettle is roughly $2\,\text{kW}$. The reactor is carrying
the heat load of several kettles continuously. At pilot scale this is
already a design constraint; at production scale it is a dominant one.
\end{estimation}

\section{Aeration, mixing, scale-up}\pathtag{core}

Aerobic biomanufacturing is usually limited by oxygen transfer before it
is limited by the solubility of glucose. Oxygen is poorly soluble in
water. A fast microbial culture can consume the dissolved oxygen in a
broth in seconds if gas transfer stops. The reactor therefore has to
move oxygen from bubbles into liquid and then from liquid to the cells
fast enough to meet demand.

\subsection{Oxygen-transfer balance}

The standard engineering expression is
\[
\text{OTR} = k_L a (C^\ast - C_L),
\]
where OTR is the oxygen-transfer rate, $k_L$ is a liquid-film
mass-transfer coefficient, $a$ is interfacial area per volume,
$C^\ast$ is the saturation concentration at the gas composition and
pressure, and $C_L$ is the dissolved oxygen concentration in the liquid.
The oxygen uptake rate is
\[
\text{OUR} = q_{O_2} X,
\]
with $q_{O_2}$ the specific oxygen-uptake rate. The oxygen setpoint is
possible only when OTR can exceed OUR with margin.

\input{volumes/06-life/ch10-bioreactors-and-biomanufacturing/figures/fig-our-otr-cellbalance.tex}

The symbols hide the engineering work. Increasing aeration raises gas
flow and often $C^\ast$; increasing agitation breaks bubbles and raises
$a$; raising pressure raises oxygen solubility; adding pure oxygen
raises the driving force. Each move has a cost: foaming, shear,
compressor power, heat, cell damage, and regulatory complexity.

\subsection{$k_L a$ correlations and estimation}

The standard empirical correlation for non-viscous stirred tanks is
Van't Riet's expression \cite{paper:v6c10-vantriet-kla-1979}:
\[
k_L a = c \left(\frac{P}{V}\right)^{a_e} v_s^{b_e},
\]
with parameters $c \approx 0.026$, $a_e \approx 0.4$, $b_e \approx 0.5$
in SI units ($k_L a$ in $\text{s}^{-1}$, $P/V$ in $\text{W}/\text{m}^3$,
$v_s$ superficial gas velocity in $\text{m}/\text{s}$). The correlation
is empirical, calibrated to water-like media, and known to overpredict
$k_L a$ in viscous filamentous broths and in foaming broths. It carries
the right scaling behaviour for engineering estimation and is the
default used in industrial scale-up exercises
\cite{paper:v6c10-junker-fermentation-scaleup-2004}.

A worked estimation. A $10\,\text{L}$ development reactor at
$P/V = 1000\,\text{W}/\text{m}^3$ and $v_s = 0.005\,\text{m}/\text{s}$
gives
\[
k_L a = 0.026 \times 1000^{0.4} \times 0.005^{0.5}
= 0.026 \times 15.8 \times 0.0707
\approx 0.029\,\text{s}^{-1} \approx 105\,\text{h}^{-1}.
\]
The order-of-magnitude $100\,\text{h}^{-1}$ matches the working data
in `data/kla_by_reactor_scale.csv`. The scaling and the operating-point
shift through scale-up are computed by `code/kla_scale_up.py` for the
three rules described next.

\input{volumes/06-life/ch10-bioreactors-and-biomanufacturing/figures/fig-kla-scale-up.tex}

\subsection{Reynolds, Sherwood, and impeller correlations}

The dimensionless framing of the gas-liquid transfer is the Sherwood
number,
\[
\text{Sh} = \frac{k_L d_b}{D_{O_2}},
\]
with $d_b$ a representative bubble diameter and $D_{O_2}$ the diffusion
coefficient of oxygen in water (about $2.1 \times 10^{-9}\,\text{m}^2/\text{s}$
at $25\,\degC$). For a bubble in a stirred medium, Sherwood is a
function of Reynolds and Schmidt:
\[
\text{Sh} = f(\text{Re}, \text{Sc}),
\quad
\text{Re} = \frac{\rho N d_i^2}{\mu},
\quad
\text{Sc} = \frac{\mu}{\rho D_{O_2}},
\]
where $N$ is impeller speed and $d_i$ impeller diameter. The impeller
Reynolds is a fluid-mechanics statement; the Schmidt is a property
statement; their combination through a correlation like
$\text{Sh} \sim \text{Re}^{0.5}\text{Sc}^{1/3}$ for surface-renewal-
governed transfer recovers a $k_L$ that scales weakly with agitation.
The bubble interfacial area $a$ is what scales strongly with power
input, which is why $k_L a$ is dominated by $a$ rather than by $k_L$ in
stirred tanks. The dimensionless framing connects the empirical Van't
Riet correlation to the underlying transport principles in Vol IV.

The impeller power for an ungassed Newtonian liquid is
\[
P_0 = N_p \rho N^3 d_i^5,
\]
with $N_p$ the impeller power number ($\sim 5$ for a six-blade Rushton
turbine at fully turbulent Reynolds). Gassing reduces the power draw
because gas bubbles lower the average broth density seen by the
impeller; the gassed-to-ungassed power ratio $P_g / P_0$ is itself a
function of gas flow and impeller geometry, and the operating-point
calculation must use $P_g$, not $P_0$, in the Van't Riet correlation.

\subsection{Mixing time and gradients}

Laboratory reactors are often close to homogeneous. Large reactors are
not. A $2\,\text{L}$ stirred tank can mix in seconds; a
$100\,\text{m}^3$ vessel may contain substrate-rich zones near the feed,
oxygen-poor zones far from spargers, pH gradients around base addition,
and thermal gradients around jackets or coils. Cells that circulate
through those zones experience a periodic environment. The process
model that uses one $S(t)$, one $C_L(t)$, and one pH is an average over
that circulation.

A useful order-of-magnitude scale for the mixing time in a stirred tank
is
\[
t_m \approx \frac{5.4}{N}\left(\frac{V}{d_i^3}\right)^{2/3} N_p^{-1/3},
\]
which at constant $P/V$ scales as $V^{2/9}$. Mixing time therefore
worsens slowly with scale at constant $P/V$ and worsens fast at
constant tip speed. A circulation time of a few seconds in a $2\,\text{L}$
flask becomes tens of seconds in a $20\,\text{m}^3$ tank; in a
$200\,\text{m}^3$ tank, regions far from the feed can see substrate
spike and oxygen drop on the order of a circulation period, which for
many \emph{E.\ coli} processes drives overflow metabolism (acetate) and
yield loss.

Scale-up therefore asks which average must be preserved. Constant
geometric similarity preserves shape. Constant power per volume
preserves one measure of mixing intensity. Constant tip speed protects
shear-sensitive cells. Constant $k_L a$ protects oxygen transfer.
Constant mixing time protects homogeneity. The list cannot be satisfied
simultaneously across a large scale change
\cite[ch.~10]{text:v6c10-doran-bioprocess-2013}.

\begin{principle}[Scale-up preserves a chosen failure margin]
A scale-up rule is a decision about which failure mode receives the
largest margin. Constant $k_L a$ says oxygen limitation is the feared
failure. Constant tip speed says shear damage is the feared failure.
Constant power per volume says mixing and suspension are the feared
failure. The engineer states the feared failure explicitly, then accepts
the secondary consequences.
\end{principle}

\subsection{Foam, shear, and gas handling}

Gas makes foam. Proteins, surfactants, cell debris, and extracellular
polymer stabilise the foam. Foam carries cells and product into filters,
vents, and exhaust lines, where it becomes contamination risk and
product loss. Antifoam reduces foam but can reduce oxygen transfer by
disrupting the bubble interface and can complicate downstream
purification \cite{paper:v6c10-bell-fermenter-foam-2014}. A foam probe
and antifoam pump are therefore part of the mass-transfer design, not
an accessory.

Shear is equally context-dependent. Bacteria tolerate high agitation.
Filamentous fungi change morphology under shear. Mammalian cells are
larger, membrane-bound, and more shear-sensitive, especially at gas-liquid
interfaces. Protecting mammalian cells often means lower agitation,
gentler sparging, microcarriers, cell-retention devices, or perfusion
formats. The reactor is chosen around the cell's mechanical tolerance.

\section{Sterile process design}\pathtag{standard}

Sterility in biomanufacturing is a process property. It is not a
sentence written after a vessel has been cleaned. A sterile process is
designed so that every material entering the boundary has a defined
sterilisation route, every connection has a defined aseptic procedure,
and every excursion has a defined response.

\subsection{The sterile boundary}

The sterile boundary includes the reactor vessel, inlet filters,
medium lines, gas lines after sterilising filters, harvest lines,
sampling ports, probes, seals, valves, and any disposable assemblies.
The boundary is drawn before the run. If a sampling operation crosses
the boundary, the sampling operation is part of the process design.

Steam-in-place and clean-in-place systems sterilise or clean hard-piped
equipment. Disposable bags and tubing sets arrive gamma-irradiated or
otherwise sterilised by the supplier, then are connected by sterile
welds, aseptic connectors, or closed manifolds. Neither route removes
the need for procedure. A sterile component handled through an open
connection in an uncontrolled environment is no longer part of a sterile
process.

The 2009 Genzyme Allston contamination, reviewed below in
\S\ref{sec:vol06ch10:failure}, exercised the same boundary at the
raw-material end. A serum-borne calicivirus
\cite{paper:v6c10-vesivirus-2117-victoria-2006} entered the upstream
through animal-derived medium components rather than through an open
process operation. The sterile boundary therefore extends through the
supply chain into the supplier's vessels, not only through the
manufacturer's plant.

\subsection{Kill kinetics and validation}

Microbial death under heat treatment is commonly approximated as
first-order on a logarithmic scale. If $D$ is the decimal reduction
time, then
\[
N(t) = N_0 10^{-t/D}.
\]
A six-log reduction reduces $10^6$ organisms to an expected count of
one. A twelve-log reduction reduces $10^{12}$ to one, in expectation.
The calculation is not a guarantee that the reactor is sterile. It is
the quantitative core of a sterilisation design that also requires cold
spot identification, biological indicators, material compatibility, and
validated hold times \cite[ch.~7]{text:stanbury-fermentation}.

\input{volumes/06-life/ch10-bioreactors-and-biomanufacturing/figures/fig-sterile-log-reduction.tex}

\textbf{A 6-log reduction worked.} Take \emph{Bacillus subtilis} spores
under saturated steam at $121\,\degC$ with $D_{121} = 2.1\,\text{min}$.
The hold time for a 6-log reduction is
\[
t_{6D} = 6 \times 2.1\,\text{min} = 12.6\,\text{min}.
\]
A 12-log target for terminal sterilisation requires $25.2\,\text{min}$.
A cold spot at $117\,\degC$ (a typical heat-up gradient at the deepest
point of a packed bioreactor or a tubing manifold) has
\[
D_{117} = D_{121} \times 10^{(121 - 117)/z}
= 2.1 \times 10^{4/10}
= 5.3\,\text{min},
\]
with $z = 10\,\degC$ the Arrhenius-like temperature dependence. The
cold spot needs $2.5\,\times$ the hold time of the setpoint for the
same log reduction. The validation programme identifies and instruments
the cold spot, biological indicators (heat-resistant spore strips of
\emph{G.\ stearothermophilus}) confirm the kill in routine
qualification runs, and the SOP carries the longer hold time. The
`sterile_log_reduction.py' script in `code/` computes the same numbers
for arbitrary temperature, $D$, and $z$.

\subsection{Sterile filtration}

Sterile filtration of liquid media uses $0.22\,\mu\text{m}$ or
$0.1\,\mu\text{m}$ membrane filters, validated to retain the standard
challenge organisms \emph{Brevundimonas diminuta} (for
$0.22\,\mu\text{m}$) and \emph{Acholeplasma laidlawii} (mycoplasma, for
$0.1\,\mu\text{m}$). The filter is part of the sterile boundary; its
integrity is a process variable. Pre-use and post-use integrity tests
(bubble point, diffusion flow, water intrusion) detect breached filters
and the result of the post-use test is recorded against the batch. A
filter that passes pre-use and fails post-use renders the downstream
material at risk and triggers a deviation; depending on the process and
the analytical evidence, the affected batch is reworked, segregated, or
rejected.

\subsection{GMP as engineering discipline}

Good manufacturing practice is often taught as paperwork. The better
engineering reading is that GMP is a memory and control system for
processes whose failures may not be visible at the end of the run. Batch
records, change control, deviation reports, validated cleaning, operator
training, and material traceability exist because a contaminated or
misprocessed biological product may look normal while carrying a hidden
hazard. The quality system is a control loop around a process whose
state cannot be fully observed in real time.

This chapter does not train the reader as a GMP operator. It gives the
engineering reason for the discipline: in a biological process, the
product's history is part of the product.

\section{Cell-culture vs microbial vs plant systems}\pathtag{standard}
\label{sec:vol06ch10:platforms}

The host is a manufacturing platform. The choice among bacteria, yeast,
filamentous fungi, mammalian cells, plant cells, plants, insects, and
cell-free systems is a process decision as much as a biological one.

\engterm{Bacterial systems}. \emph{E.\ coli} grows quickly, reaches high
cell densities, is genetically tractable, and is cheap to feed. It is a
strong platform for enzymes, peptides, non-glycosylated proteins,
metabolites, and pathway-engineered small molecules. Its weaknesses are
equally engineering-relevant: inclusion bodies, endotoxin, limited
post-translational modification, codon and folding problems, and strong
selection against burdensome products. The first commercial recombinant
biopharmaceutical, Eli Lilly's Humulin (recombinant human insulin),
was an \emph{E.\ coli} process approved by FDA in October 1982 under
license from Genentech and based on the Goeddel et al.\ 1979 expression
work \cite{paper:v6c10-goeddel-insulin-1979,
paper:v6c10-humulin-eli-lilly-1982}. Recombinant somatropin (human
growth hormone), interferons, and several enzyme-replacement therapies
followed on the same \emph{E.\ coli} platform during the 1980s and
1990s.

\engterm{Yeast and fungi}. \emph{Saccharomyces cerevisiae},
\emph{Pichia pastoris}, and filamentous fungi sit between bacteria and
mammalian cells. They are eukaryotic, tolerate industrial fermentation,
secrete many products, and can perform some post-translational
processing. They also bring glycosylation patterns that may differ from
human patterns, morphology problems in filamentous growth, and oxygen
and mixing demands that can dominate scale-up
\cite{paper:v6c10-stockmann-pichia-2009}. The second-generation
recombinant insulin programmes, including Novo Nordisk's Novolin and
its derivatives, used a yeast secretion route to produce a proinsulin
intermediate that avoided the \emph{E.\ coli} inclusion-body refolding
step.

\engterm{Mammalian cells}. CHO and related mammalian-cell platforms are
used for many therapeutic proteins because they fold, assemble, and
glycosylate proteins in ways acceptable for human therapy. They grow
slowly, require expensive media, are shear-sensitive, and demand tight
control of metabolites such as lactate and ammonia. A mammalian-cell
process is often justified by product quality rather than by ease of
manufacture. The dominant industrial mAb process is a CHO fed-batch in
a stirred-tank reactor at scales from 200 L (clinical, single-use) to
20000 L (commercial, hard-piped); titers improved from about
$0.05\,\text{g}/\text{L}$ in the 1986 cohort to $3$--$6\,\text{g}/\text{L}$
in mid-2010s commercial processes through a combination of cell-line
engineering, media optimisation, and feed-strategy work
\cite{paper:v6c10-li-cho-cellline-2010}.

\engterm{Plant and algal systems}. Plant cells, whole plants, and algae
offer photosynthetic or plant-specific production routes, large surface
area, and some low-cost cultivation possibilities. Their engineering
constraints are light delivery, slow growth, seasonal or batch
variability, containment, extraction, and product heterogeneity. A
photobioreactor is a light-transfer device before it is a tank. The
small approved-product portfolio for plant production includes
taliglucerase alfa (Elelyso, Protalix), an enzyme-replacement therapy
for Gaucher disease produced in carrot cells, which entered the same
clinical space as Genzyme Cerezyme and offered an alternative supply
during the 2009-2010 Cerezyme shortage discussed in
\S\ref{sec:vol06ch10:failure}.

\engterm{Cell-free systems}. A cell-free system uses extracted or
reconstituted biochemical machinery outside a living cell. It removes
the burden of keeping a whole organism alive and can simplify toxic
products or rapid prototyping. It also replaces cellular maintenance
with expensive reagents, short reaction lifetimes, and a different
scale-up problem. Cell-free synthesis is best read as a specialised
manufacturing route, not as a universal replacement for cells.

\section{Downstream processing}\pathtag{standard}

The reactor makes a broth, not a finished product. Downstream processing
turns that broth into material with a defined identity, purity, potency,
stability, and container. The train may include cell removal, cell
disruption, clarification, capture chromatography, precipitation,
filtration, polishing chromatography, viral clearance, concentration,
buffer exchange, formulation, sterile filtration, filling, inspection,
and storage \cite[ch.~14]{text:v6c10-shuler-bioprocess-2017,
paper:v6c10-ladisch-bioseparations-2001}.

\input{volumes/06-life/ch10-bioreactors-and-biomanufacturing/figures/fig-upstream-downstream-flowsheet.tex}

\subsection{Intracellular and extracellular product}

An extracellular product leaves the cell and appears in the medium. The
first separation removes cells from liquid. An intracellular product
requires harvesting cells and breaking them open, which releases DNA,
host-cell proteins, lipids, endotoxin in Gram-negative bacteria, and
proteases. The difference changes the whole plant. A secreted product
can be dilute but cleaner. An intracellular product can be concentrated
but dirty.

\subsection{Harvest and capture}

Harvest separates cells from liquid. Continuous centrifugation handles
microbial broths at scale; depth filtration handles mammalian-cell
broths and clarifies a broth gentler on cell fragments. For an
intracellular product, the cells are concentrated, washed, and lysed
by high-pressure homogenisation (typical operating pressure
600--1200\,\text{bar}) or by enzymatic / detergent disruption.

Capture chromatography is the first chromatographic step. Its job is
not maximum purity but maximum throughput: catch the product, drop the
volume, and remove the worst impurities (DNA, host-cell proteins,
endotoxin). The dominant capture chemistries are ion exchange (IEX,
strong cation or strong anion), Protein A affinity (for IgG monoclonal
antibodies), and mixed-mode resins. Capture step yields range from
about 0.80 (mid-resolution IEX) to about 0.95 (high-affinity Protein A).
The `chromatography_resolution.py` script in `code/` computes peak
resolution $R_s$ from retention times and peak widths and reports the
cumulative downstream yield through a configurable train.

\subsection{Polishing and viral clearance}

Intermediate and polishing chromatography (often a different
selectivity, such as hydrophobic interaction or reversed-phase after
an IEX capture) remove the impurity classes the capture step did not.
Viral clearance is an orthogonal concern for any product made in a
mammalian-cell or animal-derived raw-material process. The standard
viral clearance toolkit combines a low-pH inactivation hold (for
enveloped viruses), a $20\,\text{nm}$ nanofilter (for small non-
enveloped viruses), and chromatographic clearance (additional log
reduction during routine polishing). The regulatory requirement is to
demonstrate the train's clearance capacity for a panel of model
viruses, typically achieving an aggregate $>15$ log reduction against
each model. The 2009 Genzyme contamination event exposed the same
calculation: the downstream train cleared Vesivirus 2117 to a degree
that satisfied a post-event review, but the regulatory requirement
moved to virus-free media and tighter raw-material control rather
than relying on downstream clearance alone
\cite{web:v6c10-fda-warning-letter-genzyme-2010}.

\subsection{Yield through the train}

Downstream recovery is multiplicative. A train with five steps at 90\%
yield each has total yield
\[
0.9^5 = 0.59.
\]
At 80\% each, it has
\[
0.8^5 = 0.33.
\]
That arithmetic is why an apparently small improvement in a capture
step can be more valuable than a larger increase in reactor titer. The
process engineer asks where each gram is lost, and whether the lost gram
is product, impurity, or material that protects the product from damage.

\subsection{Purity, potency, and comparability}

For a therapeutic protein, the product is not only the amino-acid
sequence. Glycosylation, folding state, aggregation, clipping, oxidation,
host-cell protein profile, DNA carryover, endotoxin, and formulation
state can affect safety and efficacy. The process defines the product.
Changing the host, media, reactor scale, purification resin, hold time,
or formulation can change the product enough that comparability becomes
an engineering and regulatory question
\cite{paper:v6c10-jagschies-biopharma-economics-2018}.

\section{Worked example: insulin manufacture}\pathtag{standard}
\label{sec:vol06ch10:insulin}

Recombinant human insulin is a useful worked example because it connects
the molecular design of Chapter~\ref{vol06:ch09} to the process design
of this chapter. Early recombinant insulin work expressed chemically
synthesised human-insulin genes in \emph{E.\ coli}, with the A-chain and
B-chain produced separately and joined chemically
\cite{paper:v6c10-goeddel-insulin-1979}. Eli Lilly's Humulin, marketed
under license from Genentech, received FDA approval in October 1982 and
became the first commercial recombinant biopharmaceutical
\cite{paper:v6c10-humulin-eli-lilly-1982,
paper:v6c10-rader-biopharma-stats-2008}. Second-generation processes
moved to a single proinsulin expression with an internal cleavage step
(some in \emph{E.\ coli}, some in yeast secretion). Modern processes
vary by producer and product, but the engineering structure is stable:
construct a production strain, grow it under controlled conditions,
express an insulin precursor, recover and purify the protein, process
it to the active form, polish impurities, formulate, and fill.

\input{volumes/06-life/ch10-bioreactors-and-biomanufacturing/figures/fig-insulin-process.tex}

We will not claim a current commercial process. We will design a toy
process whose numbers are explicit and reviewable.

\subsection{A toy mass balance}

Suppose the target is $1.0\,\text{kg}$ of purified insulin-equivalent
product. Assume:
\[
\text{final recovery from culture to purified product} = 0.40,
\]
and an achieved product titer in culture of
\[
1.0\,\text{g}/\text{L}.
\]
The required culture volume is
\[
V = \frac{1000\,\text{g}}{0.40 \times 1.0\,\text{g}/\text{L}}
= 2500\,\text{L}.
\]
If the plant uses $500\,\text{L}$ production batches, five successful
batches are required. If the batch rejection rate is 10\%, the expected
number of batches attempted is $5/0.9 \approx 5.6$. The half-batch in
that arithmetic is not a real batch; it is the planning signal that a
campaign of five batches has poor margin.

Now include downstream yield. Suppose the train has four major steps:
cell harvest or clarification, capture, conversion or refolding, and
polishing/formulation. If their yields are 0.90, 0.80, 0.75, and 0.75,
then total yield is
\[
0.90 \times 0.80 \times 0.75 \times 0.75 = 0.405.
\]
The assumed 40\% recovery is therefore not a decorative parameter. It is
the product of four specific engineering performances. The
`insulin_yield_calc.py' script runs the same calculation with
sensitivity sweeps for titer and weakest-step yield.

\subsection{From E.\ coli inclusion body to active insulin}

A representative early commercial route, simplified for the toy
calculation, has the following stages:

\begin{enumerate}[leftmargin=*]
\item \textbf{Strain construction.} An expression vector carrying a
  fusion-protein gene (insulin chain plus a fusion partner that
  partitions the product into an inclusion body) is transformed into a
  production \emph{E.\ coli} strain.
\item \textbf{Inoculum train.} A working-cell-bank vial is grown
  through a seed train (shake flask, seed reactor) into a production
  bioreactor. Working volumes 10 to 50 m$^3$ at full commercial scale.
\item \textbf{Fed-batch fermentation.} Production runs at an
  exponential feed designed to hold growth substrate near
  $K_S$-multiple of the strain so that overflow metabolism is avoided
  and oxygen demand stays within the reactor's $k_L a$. Induction is
  often by IPTG or by metabolite shift; the product accumulates in
  inclusion bodies. Endpoint OD$_{600}$ on the order of 100 in modern
  high-cell-density fed-batches.
\item \textbf{Cell harvest.} Continuous disc-stack centrifugation
  concentrates the cells to a slurry of about 25--30\% solids by mass.
\item \textbf{Cell lysis.} High-pressure homogenisation at
  600--1000\,\text{bar} disrupts the cells in a single or double pass.
  The released debris is removed by centrifugation and inclusion bodies
  are washed.
\item \textbf{Solubilisation.} Inclusion bodies are dissolved in a
  denaturant (urea or guanidinium chloride) with a reducing agent
  (DTT or beta-mercaptoethanol) to break the misfolded disulfides.
\item \textbf{Oxidative refolding.} The denaturant is diluted; oxidation
  conditions (controlled redox couple) are imposed; the chain refolds
  to the native two-chain disulfide pattern at low protein
  concentration (typically $0.1$--$0.5\,\text{g}/\text{L}$). Refolding
  yield 60--80\% is typical.
\item \textbf{Enzymatic cleavage.} For a proinsulin or fusion-protein
  route, an enzymatic cleavage step liberates the mature two-chain
  insulin.
\item \textbf{Capture and polishing.} Ion-exchange chromatography
  captures the product. Reversed-phase chromatography polishes the
  preparation and removes related-impurity peaks (oxidised forms,
  des-amido variants, fragments).
\item \textbf{Size-exclusion polishing.} A final size-exclusion step
  removes aggregates.
\item \textbf{Crystallisation.} Zinc-insulin crystallisation produces a
  stable bulk intermediate.
\item \textbf{Formulation and fill.} The bulk is dissolved into the
  finished-product buffer (phosphate or tris, zinc, preservative),
  sterile-filtered, and filled into vials or cartridges.
\end{enumerate}

The whole train is the figure
\cref{fig:vol06ch10:insulin-process}. The economics carried by the
train are documented in standard industry references; the current
estimated annual world insulin demand is on the order of
$10\,000\,\text{kg}$ active insulin equivalent
\cite{paper:v6c10-basu-insulin-access-2019}, and the production
footprint that produces it consists of fewer than a dozen large plants
operated by three principal manufacturers (Eli Lilly, Novo Nordisk,
Sanofi).

\subsection{What the example teaches}

The reactor titer matters, but the process does not end at harvest.
Double the titer and hold downstream recovery fixed, and the required
culture volume halves. Improve the weakest downstream step from 75\% to
90\%, and total recovery rises from 40.5\% to 48.6\%, a 20\% gain in
purified output at the same reactor titer. Reduce batch rejection from
10\% to 2\%, and campaign planning changes again. The process optimum
is distributed across strain, reactor, purification train, quality
system, and scheduling.

The example also shows why biomanufacturing belongs in Volume VI rather
than only in Volume VIII or XI. The machine is the cell plus the vessel
plus the purification train plus the quality system. The living phase is
part of the machine.

\begin{estimation}
\textbf{Estimate the worldwide annual insulin demand and the
fermenter capacity required to supply it.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The world has about $8.0$ billion people (current as of 2024); the
diabetes prevalence is about $9.3\,\%$, so about $750$ million people
have diabetes; the fraction requiring insulin (type 1 plus
insulin-dependent type 2) is roughly $10\,\%$, giving about
$75$ million insulin-using patients.

A typical dose is about $40\,\text{units}/\text{day}$, where one
international unit (IU) of insulin corresponds to about
$0.0347\,\text{mg}$ of active hormone. The per-patient daily mass is
\[
40 \times 0.0347 \,\text{mg}/\text{day}
\approx 1.4\,\text{mg}/\text{day}
\approx 0.5\,\text{g}/\text{year}.
\]
The world annual demand is therefore
\[
75 \times 10^6 \times 0.5\,\text{g}/\text{year}
\approx 3.7 \times 10^7\,\text{g}/\text{year}
\approx 3.7 \times 10^4\,\text{kg}/\text{year}
\approx 37\,\text{tonnes}/\text{year}.
\]
The peer-reviewed Basu et al.\ 2019 microsimulation reports a global
2018 figure of about $516\,\text{million}$ vials of $1000$ IU, which
back-translates to roughly the same order
\cite{paper:v6c10-basu-insulin-access-2019}; the working-data table at
`data/insulin_demand_by_region.csv' gives a region-by-region breakdown
totalling about $8.6$ tonnes for the lowest-end estimate, and the
divergence between the two figures captures the access gap in
low-income regions.

At an industrial fermentation titer of $1\,\text{g}/\text{L}$ and a
downstream recovery of 40\%, $37\,\text{tonnes}/\text{year}$ requires
\[
\frac{37\,000\,\text{kg}/\text{year}}{0.40 \times 1\,\text{g}/\text{L}}
= 9.3 \times 10^7\,\text{L}/\text{year},
\]
that is, about $93\,000\,\text{m}^3$ of culture per year. A
single $50\,\text{m}^3$ fed-batch fermenter at $250\,\text{h}$ per
batch (10 days for inoculation, growth, induction, harvest, and
turnaround) and an availability of 80\% delivers about
\[
\frac{50\,\text{m}^3 \times 8000\,\text{h}/\text{year}}{250\,\text{h}/\text{batch}}
\times 0.80 \approx 1280\,\text{m}^3/\text{year},
\]
so the world demand fits in about $70$ to $100$ such fermenters
running in parallel. That number matches the small-double-digit
count of large insulin fermenters operated worldwide by the three
principal manufacturers, with the geographic concentration in
Indianapolis (Lilly), Kalundborg (Novo Nordisk), and Frankfurt
(Sanofi).
\end{estimation}

\section{Failure: contamination, scale-up loss, batch failures}\pathtag{core}
\label{sec:vol06ch10:failure}

Biomanufacturing fails in ways that look familiar from the rest of
engineering, and in ways specific to living systems.

\input{volumes/06-life/ch10-bioreactors-and-biomanufacturing/figures/fig-batch-failure-tree.tex}

\engterm{Contamination}. A contaminating organism can consume substrate,
kill the production organism, release proteases, change pH, produce
toxins, or quietly introduce an impurity profile that appears only in
late analytics. In microbial production, a phage can collapse a culture
rapidly. In mammalian culture, mycoplasma or adventitious virus can
threaten a cell bank or production campaign. The response is layered:
cell-bank qualification, raw-material control, sterile boundary design,
environmental monitoring, in-process analytics, and a deviation system
that treats anomalies as information. The contamination mechanisms
worth tracking are filter integrity failures (an upstream sterilising
filter fails between pre-use and post-use integrity tests), aseptic
technique breaches (an operator-introduced organism via a connection
or sampling operation), and raw-material contamination (an organism
arrives in a medium component from the supplier).

The 2009 Genzyme Cerezyme contamination at the Allston Landing,
Massachusetts facility is the modern case study for raw-material-borne
contamination of a CHO mammalian-cell process. On 16 June 2009 Genzyme
announced detection of Vesivirus 2117 in a 2000 L CHO bioreactor
producing imiglucerase (Cerezyme) and agalsidase beta (Fabrazyme); the
contamination most plausibly entered through bovine-origin medium
components \cite{paper:v6c10-vesivirus-2117-victoria-2006}. Production
was suspended, the affected train drained and sanitised, raw-material
qualification tightened, and screening assays added. The downstream
purification trains for Cerezyme and Fabrazyme include nanofiltration
and chromatographic clearance steps validated against small non-
enveloped viruses, so patient infection from contaminated product was
not the documented risk; the documented harm was the multi-year supply
shortage that followed. The episode cost Genzyme roughly
\$300\,million in 2009 revenue, triggered a 2010 FDA Warning Letter and
a Consent Decree requiring \$175\,million in disgorgement
\cite{web:v6c10-genzyme-sec-8k-2009-06,
web:v6c10-fda-warning-letter-genzyme-2010,
web:v6c10-fda-consent-decree-genzyme-2010}, and forced patient-supply
rationing for the global Gaucher and Fabry communities
\cite{paper:v6c10-genzyme-shortage-2010}. Sanofi acquired Genzyme in
2011 for \$20.1\,billion against the backdrop of the Allston
remediation programme. The full case is summarised in the named-cases
registry at
\path{docs/research/accidents/genzyme-cerezyme-2009.md}\cite{acc:genzyme-cerezyme-2009}.
The engineering lesson runs in two directions: bovine-origin raw
materials are part of the sterile boundary; and single-site,
single-train manufacture of a sole-source biologic is a supply-chain
failure mode that contamination can exercise.

\engterm{Scale-up loss}. A process that works in a shake flask may fail
in a stirred tank because oxygen transfer, mixing time, pH control,
foam, heat removal, or shear changed. The laboratory run may have had
one environment; the production cell sees a loop through zones. The
symptom is often lower titer, altered product quality, or unstable
metabolism rather than an obvious mechanical failure. The Van't Riet
$k_L a$ correlation gives a quantitative diagnostic: a development
reactor at $P/V = 1000\,\text{W}/\text{m}^3$ yields about
$k_L a \approx 100\,\text{h}^{-1}$, and a constant-geometry scale-up to
10\,m$^3$ at the same impeller speed gives $P/V \approx 200\,\text{W}/\text{m}^3$
and $k_L a \approx 50\,\text{h}^{-1}$, a halving of oxygen-transfer
capacity. If the cell's specific oxygen-uptake rate is unchanged, the
operating point moves into oxygen limitation; if dissolved oxygen is
maintained by switching to pure oxygen, the operating cost rises and
foaming worsens. The 1986-1990 Genentech tPA early-process work, in
which yields and product-quality attributes shifted between
development scale and pilot scale, is the standard pedagogical case
for the mammalian-cell scale-up problem
\cite{paper:v6c10-li-cho-cellline-2010}.

\engterm{Genetic and phenotypic drift}. A production strain is under
selection. If product formation burdens growth, non-producing or
low-producing variants can outgrow the desired population. Plasmids can
be lost, promoters can mutate, copy number can shift, epigenetic or
phenotypic states can change, and cell-line productivity can decay. The
engineering countermeasure is to define the cell bank, passage number,
selection pressure, genetic stability test, and end-of-production
criterion. The cell-bank discipline is the same in microbial and
mammalian processes: a Master Cell Bank (MCB) is laid down once,
qualified, and stored; a Working Cell Bank (WCB) is derived from the
MCB at defined passage; production runs only from WCB vials at or
below the qualified end-of-production cell age.

\engterm{Batch rejection}. A batch can be rejected even if the reactor
looked healthy. A product-quality attribute may sit outside
specification; a filter-integrity test may fail; a temperature excursion
may invalidate a hold; a record may reveal that the material history is
not reconstructable. The rejection is expensive because the process has
already consumed media, vessel time, operator time, quality-control
capacity, and downstream resources. A rejected biological batch is a
material loss and an information loss: the plant must learn whether the
failure was a one-off excursion or a sign that the process model is
wrong. Industry data current as of mid-2020s public surveys show
overall batch-rejection rates around 8-9\% for microbial processes and
around 12-15\% for mammalian processes, distributed across the failure
modes in `data/batch_failure_modes.csv'.

\begin{principle}[The batch record is part of the product]
In biomanufacturing, a vial is acceptable only if its material history
is acceptable. The batch record, environmental record, cell-bank record,
deviation record, and analytical record are therefore part of the
engineered object. A product whose history cannot be reconstructed is
not a product the engineer can release.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Ferment a small batch and close the mass balance]
The reader performs a household-scale fermentation and treats it as a
bioreactor.

\textbf{Track}. Household fermentation plus analysis.
\textbf{Hazard class}: low. Use a standard food-safe recipe. Use clean
glassware or food-grade containers. Do not seal a gas-producing
fermentation in a rigid closed vessel. Discard the product if there is
visible mould, putrid odour, unexpected colour, gas pressure in a sealed
container, or any doubt about safety.

\textbf{Allowed systems}. Yogurt, sauerkraut, kimchi, sourdough starter,
or a sugar-yeast CO$_2$ fermentation whose product is not consumed.

\textbf{Specification}.

\begin{enumerate}[leftmargin=*]
\item Draw the process boundary. State every input mass: water, milk,
  cabbage, salt, sugar, starter, inoculum, and container tare.
\item Record temperature, time, pH if available, visible gas formation,
  mass loss if measurable, and qualitative state at least once per day.
\item Write a mass balance for the run. For a CO$_2$-producing
  fermentation, estimate the CO$_2$ produced from the mass loss or from
  stoichiometry. For yogurt or lactic fermentation, estimate acid
  production from pH or from literature values and state the uncertainty.
\item Identify the limiting substrate, expected product, expected waste,
  and principal contamination routes.
\item Compare the observed time course to a simple growth or conversion
  model. State which parameter you would need to measure to improve the
  model.
\end{enumerate}

\textbf{Tools}. The `code/mass_balance_ferment.py' script in this
chapter computes the salt fraction, brine, lactate, and CO$_2$ output
for a sauerkraut batch, and the ethanol and CO$_2$ output for a
sugar-yeast batch, given the input masses. Use it to bracket the
expected mass loss in advance and compare to your kitchen-scale data.

\textbf{Deliverable}. A 6- to 10-page engineering notebook report with
the boundary drawing, raw data table, mass balance, uncertainty
estimate, failure-mode discussion, and a one-page reflection on which
parts of the living process were observable and which were hidden.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A batch culture starts with $S_0 = 20\,\text{g}/\text{L}$ glucose and
has biomass yield $Y_{X/S} = 0.45\,\text{g}/\text{g}$. Ignore
maintenance and product formation. Estimate the maximum biomass
concentration increase.
\end{exercise}

\paragraph{Solution sketch.}
At complete consumption, $\Delta X = Y_{X/S}\,\Delta S = 0.45 \times
20 = 9\,\text{g}/\text{L}$ biomass produced. The maximum biomass
concentration at harvest is therefore $X_0 + 9\,\text{g}/\text{L}$;
for a typical inoculum of $X_0 \approx 0.1\,\text{g}/\text{L}$, the
final biomass concentration is about $9\,\text{g}/\text{L}$. The
ignored maintenance term $m_S X$ would shave this down by perhaps
10-20\% for a slow microbial process; ignoring product formation is
acceptable if the product is small or co-formed with biomass, and is
not when the product is a large fraction of the substrate (lactic
acid at $0.95\,\text{g}/\text{g}$ would invalidate the calculation).

\begin{exercise}
A chemostat has volume $2.0\,\text{L}$ and feed flow
$0.60\,\text{L}/\text{h}$. Compute the dilution rate. If
$\mu_{\max}=0.50\,\text{h}^{-1}$, is washout possible at saturating
substrate?
\end{exercise}

\paragraph{Solution sketch.}
$D = F/V = 0.60/2.0 = 0.30\,\text{h}^{-1}$. The washout condition is
$D > \mu_{\max}$. Here $D = 0.30 < \mu_{\max} = 0.50$, so steady
operation is possible at sufficiently high feed substrate. The
steady-state substrate is
$S^\ast = K_S D / (\mu_{\max} - D) = K_S \cdot 0.30 / 0.20 = 1.5 K_S$,
and the steady-state biomass is $X^\ast = Y_{X/S}(S_f - S^\ast)$. If
operating-condition perturbations (pH drift, temperature excursion,
phage burden) reduce $\mu_{\max}$ to below $0.30\,\text{h}^{-1}$, the
same dilution rate becomes washout: the culture dilutes faster than
it grows.

\begin{exercise}
For Monod parameters $\mu_{\max}=0.8\,\text{h}^{-1}$ and
$K_S=0.10\,\text{g}/\text{L}$, compute the growth rate at
$S=0.05$, $0.10$, $1.0$, and $10\,\text{g}/\text{L}$.
\end{exercise}

\paragraph{Solution sketch.}
$\mu = 0.8 S / (0.10 + S)$, so:
\[
\mu(0.05) = 0.8 \times 0.05/0.15 = 0.267\,\text{h}^{-1},
\]
\[
\mu(0.10) = 0.8 \times 0.10/0.20 = 0.40\,\text{h}^{-1} \;(= \mu_{\max}/2),
\]
\[
\mu(1.0)  = 0.8 \times 1.0/1.1   = 0.727\,\text{h}^{-1},
\]
\[
\mu(10)   = 0.8 \times 10/10.1   = 0.792\,\text{h}^{-1}.
\]
The growth rate at $S = K_S$ is exactly $\mu_{\max}/2$, by construction.
At $S = 10 K_S$ the rate is within 10\% of $\mu_{\max}$; at
$S = 100 K_S$ the rate is within 1\%. This is the reason batch culture
spends most of its early course at near-$\mu_{\max}$ growth.

\begin{exercise}
A reactor has $k_La = 180\,\text{h}^{-1}$, $C^\ast =
7.0\,\text{mg}/\text{L}$, and a dissolved-oxygen setpoint
$C_L = 2.0\,\text{mg}/\text{L}$. Compute the oxygen-transfer rate in
$\text{mg}/(\text{L}\cdot\text{h})$.
\end{exercise}

\paragraph{Solution sketch.}
$\text{OTR} = k_L a (C^\ast - C_L)
= 180 \times (7.0 - 2.0)
= 180 \times 5.0
= 900\,\text{mg}/(\text{L}\cdot\text{h})
= 0.028\,\text{mol O}_2/(\text{L}\cdot\text{h})$.
This OTR can support a specific oxygen-uptake rate of about
$2.8\,\text{mmol O}_2/(\text{g}\cdot\text{h})$ at $X = 10\,\text{g}/\text{L}$,
which is the upper edge of typical \emph{E.\ coli} aerobic demand.
Operating with no margin is risky; design at $C_L \approx 1.5
C_{\text{setpoint}}$ or raise $k_L a$ to about
$300\,\text{h}^{-1}$ to keep OTR roughly 50\% above peak OUR.

\begin{exercise}
A $500\,\text{L}$ culture produces $0.8\,\text{g}/\text{L}$ product at
harvest. Downstream recovery is 35\%. Compute purified product per
batch and the number of successful batches needed for $1.0\,\text{kg}$.
\end{exercise}

\paragraph{Solution sketch.}
Product in broth: $500\,\text{L} \times 0.8\,\text{g}/\text{L} =
400\,\text{g}$ per batch. Purified product per successful batch:
$400 \times 0.35 = 140\,\text{g}$. Number of successful batches for
$1000\,\text{g}$: $1000/140 \approx 7.1$. At a 10\% rejection rate,
expected batches attempted: $7.1 / 0.9 \approx 7.9$. The campaign is
therefore eight planned batches with a tight buffer; doubling either
titer or recovery halves the count and the calendar.

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the nonzero steady-state biomass and substrate concentrations
for a chemostat with Monod growth and biomass yield $Y_{X/S}$.
\end{exercise}

\paragraph{Solution sketch.}
At steady state, $dX/dt = 0$ and $dS/dt = 0$. From the biomass balance,
$(\mu(S) - D) X^\ast = 0$, so either $X^\ast = 0$ (washout) or
$\mu(S^\ast) = D$. The Monod expression gives
\[
\mu_{\max} \frac{S^\ast}{K_S + S^\ast} = D
\quad\Rightarrow\quad
S^\ast(\mu_{\max} - D) = D K_S
\quad\Rightarrow\quad
S^\ast = \frac{K_S D}{\mu_{\max} - D}.
\]
The substrate balance at steady state is
$D(S_f - S^\ast) = (1/Y_{X/S})\mu(S^\ast) X^\ast$, and using
$\mu(S^\ast) = D$ gives
\[
X^\ast = Y_{X/S}(S_f - S^\ast)
       = Y_{X/S}\left(S_f - \frac{K_S D}{\mu_{\max} - D}\right).
\]
Both are valid only for $0 < D < \mu_{\max}$ and $S^\ast < S_f$. The
productivity $D X^\ast$ has a maximum at the interior dilution rate
$D_{\text{opt}} = \mu_{\max}(1 - \sqrt{K_S/(K_S + S_f)})$, which the
reader can verify by differentiation.

\begin{exercise}
Starting from the batch equations with $\mu=\mu_{\max}$ constant and no
maintenance, derive $X(t)$ and $S(t)$ until the substrate is exhausted.
\end{exercise}

\paragraph{Solution sketch.}
With $\mu = \mu_{\max}$ constant, $dX/dt = \mu_{\max} X$ gives
\[
X(t) = X_0 \exp(\mu_{\max} t).
\]
The substrate balance is $dS/dt = -(1/Y_{X/S}) \mu_{\max} X$, so
\[
S(t) = S_0 - \frac{X(t) - X_0}{Y_{X/S}}
     = S_0 - \frac{X_0}{Y_{X/S}}\left(\exp(\mu_{\max} t) - 1\right).
\]
Substrate exhaustion at $t_e$ when $S(t_e) = 0$:
\[
\exp(\mu_{\max} t_e) = 1 + \frac{Y_{X/S} S_0}{X_0},
\quad
t_e = \frac{1}{\mu_{\max}}
\ln\!\left(1 + \frac{Y_{X/S} S_0}{X_0}\right).
\]
For $S_0 = 20\,\text{g}/\text{L}$, $X_0 = 0.1\,\text{g}/\text{L}$,
$Y_{X/S} = 0.45$, and $\mu_{\max} = 0.8\,\text{h}^{-1}$, the bracket
inside the logarithm is $1 + 90 = 91$, and $t_e = \ln 91 / 0.8
\approx 5.6\,\text{h}$. The Monod term begins to matter only in the
final half-hour as $S$ falls into the $K_S$ range.

\begin{exercise}
Derive the relation between oxygen uptake rate and heat release using
an assumed heat release of $14\,\text{MJ}/\text{kg O}_2$. Express the
answer in $\text{kW}$ for an arbitrary OUR in
$\text{mol}/(\text{L}\cdot\text{h})$ and volume in litres.
\end{exercise}

\paragraph{Solution sketch.}
Total oxygen uptake per unit time: $\text{OUR}_{\text{tot}} =
\text{OUR}\cdot V$ in mol/h. Mass of O$_2$ per hour:
$M_{O_2} = 0.032\,\text{kg/mol} \cdot \text{OUR}\cdot V$, so
$M_{O_2}$ in kg/h. Heat release:
\[
\dot Q = 14\,\frac{\text{MJ}}{\text{kg}} \times M_{O_2}
       = 14 \times 0.032 \times \text{OUR}\cdot V
       = 0.448\,\text{MJ/h}\,\cdot\,\text{OUR}\cdot V.
\]
Converting MJ/h to kW (divide by 3.6): $\dot Q$ in kW =
$0.448/3.6\,\cdot\,\text{OUR}\cdot V \approx 0.124\,\text{OUR}\cdot V$,
with OUR in $\text{mol}/(\text{L}\cdot\text{h})$ and $V$ in litres.
Apply to the chapter's worked case: $V = 1000\,\text{L}$,
$\text{OUR} = 0.10\,\text{mol}/(\text{L}\cdot\text{h})$ gives
$\dot Q \approx 12.4\,\text{kW}$, matching the estimation block.

\begin{exercise}
For a sterilisation process with decimal reduction time $D$, derive the
time required for an $n$-log reduction. Apply it to $D=2\,\text{min}$
and $n=12$.
\end{exercise}

\paragraph{Solution sketch.}
From $N(t) = N_0 10^{-t/D}$, an $n$-log reduction requires
$N(t)/N_0 = 10^{-n}$, so $t/D = n$ and $t_{nD} = n D$. At $D = 2\,\text{min}$
and $n = 12$, $t_{12D} = 24\,\text{min}$. Note that an
expected count of $10^{-12} N_0$ is not zero; it is a probability
statement about a single starting load. A real validation programme
verifies the kill by biological indicator (\emph{G.\ stearothermophilus}
spore strips placed at the most-difficult-to-sterilise positions),
not by the logarithmic calculation alone.

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the time for a $1\,\text{L}$ well-mixed culture at
$X=5\,\text{g}/\text{L}$ to consume the dissolved oxygen if aeration
stops. State your assumed oxygen solubility and specific uptake rate.
\end{exercise}

\paragraph{Solution sketch.}
Assume $C^\ast = 7.0\,\text{mg O}_2/\text{L}$ at the working
temperature ($30\,\degC$, air). Initial dissolved oxygen mass at full
saturation: $7.0\,\text{mg}$ in the litre. Assume specific oxygen
uptake $q_{O_2} = 10\,\text{mmol O}_2/(\text{g}\cdot\text{h})$ at full
metabolic activity (typical for fast-growing \emph{E.\ coli}). The
oxygen-uptake rate per litre is
\[
\text{OUR} = q_{O_2} X
= 10\,\text{mmol}/(\text{g}\cdot\text{h})\times 5\,\text{g/L}
= 50\,\text{mmol}/(\text{L}\cdot\text{h})
= 1600\,\text{mg O}_2/(\text{L}\cdot\text{h}).
\]
Time to consume $7.0\,\text{mg}$ at $1600\,\text{mg}/\text{h}$:
\[
t \approx \frac{7.0}{1600} \times 3600\,\text{s} \approx 16\,\text{s}.
\]
The culture is therefore seconds from oxygen exhaustion if aeration
stops. The implication is that the dissolved-oxygen probe has to
respond in seconds, not minutes; that an interruption in agitation or
gas flow shows up first as a fast DO drop; and that anaerobic excursions
in a nominally aerobic process do not require long failures.

\begin{exercise}
Estimate the heat load of a $10{,}000\,\text{L}$ aerobic reactor whose
oxygen uptake is $0.05\,\text{mol}/(\text{L}\cdot\text{h})$.
\end{exercise}

\paragraph{Solution sketch.}
Using the derivation in Exercise 8: $\dot Q \approx 0.124 \cdot
\text{OUR} \cdot V$ kW. Here OUR $= 0.05\,\text{mol}/(\text{L}\cdot\text{h})$
and $V = 10\,000\,\text{L}$, so $\dot Q \approx 0.124 \times 0.05
\times 10\,000 \approx 62\,\text{kW}$. The reactor must therefore
reject the heat of about $30$ kettles continuously through its jacket
or coil. The cooling water flow is set by $\dot Q = m_c C_p \Delta T$;
at $\Delta T = 10\,\degC$ between inlet and outlet,
$m_c = 62\,000 / (4.18 \times 10) \approx 1500\,\text{kg/h}$ of
cooling water. At larger scale or higher OUR the cooling design
becomes a dominant capital expense.

\begin{exercise}
Estimate how many $2000\,\text{L}$ batches are needed to produce
$10\,\text{kg}$ purified product if the culture titer is
$2\,\text{g}/\text{L}$ and total recovery is 50\%.
\end{exercise}

\paragraph{Solution sketch.}
Product in broth per batch: $2000\,\text{L} \times 2\,\text{g}/\text{L}
= 4000\,\text{g}$. Purified per batch: $4000 \times 0.5 = 2000\,\text{g}
= 2\,\text{kg}$. Successful batches for $10\,\text{kg}$:
$10/2 = 5$. At 10\% rejection, expected attempted: $5/0.9 \approx 5.6$;
at 5\% rejection, $5/0.95 \approx 5.3$. Five clean batches deliver
the campaign without margin; the planner should budget six and
schedule the sixth as the contingency slot.

\begin{exercise}
Estimate the salt concentration in a sauerkraut fermentation made from
$1.5\,\text{kg}$ cabbage and $30\,\text{g}$ salt. Express the result
as percent mass fraction and state the main uncertainty.
\end{exercise}

\paragraph{Solution sketch.}
The salt mass fraction relative to cabbage plus salt is
\[
\frac{30}{1500 + 30} = \frac{30}{1530} \approx 1.96\,\%.
\]
This sits just inside the 1.8-2.5\% Pediococcus / Leuconostoc window
that selects for lactic acid bacteria over spoilage organisms. The
main uncertainty is the cabbage's water-fraction variation (typically
90-94\%, depending on freshness): the active fermentation occurs in
the brine drawn from the cabbage, and the salt concentration in that
brine, not in the cabbage-plus-salt total, is what selects. A more
useful calculation tracks the salt in the released brine, which at
$\sim 92\%$ cabbage water gives about $30\,\text{g}/(0.92 \times
1500)\,\text{g} \approx 2.2\,\%$ salt in the brine. The
`mass_balance_ferment.py' script in `code/' does the full balance.

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Simulate a fed-batch reactor with Monod growth and a feed rate chosen to
hold substrate near $0.2\,\text{g}/\text{L}$. Plot $X$, $S$, and
reactor volume over time.
\end{exercise}

\paragraph{Discussion.}
The simulation integrates
$dX/dt = \mu X - (F/V)X$, $dS/dt = (F/V)(S_f - S) - (1/Y)\mu X$,
$dV/dt = F$, with $\mu = \mu_{\max} S / (K_S + S)$ and a control law
that adjusts $F$ to drive $S$ toward the setpoint. A simple
exponential feed profile $F(t) = F_0 \exp(\mu_{\text{set}} t)$ with
$\mu_{\text{set}} < \mu_{\max}$ holds the substrate near the setpoint
in the exponential phase; a closed-loop controller (PID on the
$S$ measurement, or on its proxy DO or CER) handles the late phase as
the reactor fills. The expected qualitative output: $X$ grows
exponentially in early phase, plateaus as $V$ fills and $F/V$ falls;
$S$ stays near the setpoint until the feed runs out; $V$ rises
asymptotically toward maximum working volume. The simulation reveals
the tradeoff between maximum biomass productivity (high $\mu_{\text{set}}$)
and overflow-metabolism risk (must keep $S$ low enough that
acetate / ethanol does not accumulate).

\begin{exercise}
Simulate a chemostat after a step increase in dilution rate from
$0.2$ to $0.6\,\text{h}^{-1}$. Use Monod kinetics and identify whether
the culture reaches a new steady state or washes out.
\end{exercise}

\paragraph{Discussion.}
With $\mu_{\max} = 0.8\,\text{h}^{-1}$ and $K_S = 0.05\,\text{g}/\text{L}$,
both $D = 0.2$ and $D = 0.6$ are below $\mu_{\max}$ and a steady state
exists at each. At $D = 0.2$, $S^\ast = 0.05 \times 0.2/0.6 = 0.017\,\text{g/L}$;
at $D = 0.6$, $S^\ast = 0.05 \times 0.6/0.2 = 0.15\,\text{g/L}$. The
biomass at the new steady state is $Y_{X/S}(S_f - S^\ast)$, which is
slightly lower at the higher dilution because $S^\ast$ is higher. The
transient passes through a damped oscillation in $S$ and $X$ that
settles on order $1/\mu_{\max}$ time scale. If instead $\mu_{\max}$
were reduced to $0.5\,\text{h}^{-1}$ (perhaps by a pH or temperature
excursion), the same step to $D = 0.6\,\text{h}^{-1}$ would produce
washout: $X$ falls exponentially with rate $D - \mu(S_f) \approx
0.6 - 0.485 = 0.12\,\text{h}^{-1}$, and the reactor would lose half
its biomass in about $5.8\,\text{h}$. The simulation makes the
washout dynamic concrete.

\begin{exercise}
Build a two-zone scale-up model in which cells circulate between an
oxygen-rich zone and an oxygen-poor zone. Compare predicted product
formation to a single well-mixed model with the same average oxygen.
\end{exercise}

\paragraph{Discussion.}
The two-zone model splits the volume into a fraction $\alpha$ near
the sparger (oxygen-rich, $C_L^A$) and $1 - \alpha$ far from the
sparger (oxygen-poor, $C_L^B$). Cells exchange between zones with a
circulation rate $\tau_c^{-1}$ (mixing-time inverse). A cell's
product-formation rate $q_P$ is a function of local $C_L$; if
$q_P(C_L)$ is non-linear (saturating or threshold-like), the volume-
averaged $\langle q_P \rangle$ differs from $q_P(\langle C_L \rangle)$.
For a saturating Monod-type $q_P$, the averaging inequality (Jensen) gives
$\langle q_P \rangle \le q_P(\langle C_L \rangle)$: the heterogeneous
reactor produces less than the homogeneous one. For an overflow-
metabolism-prone strain (acetate accumulation when $C_L$ drops), the
heterogeneous reactor also produces \emph{worse} quality, because
cells repeatedly cycle through anaerobic zones and the acetate burden
accumulates faster than in the well-mixed average. The simulation
therefore predicts both a yield loss and a quality shift relative to
the well-mixed prediction, and the magnitude grows with the mixing-
time ratio $\tau_c / (1/q_{O_2})$.

\begin{exercise}
Monte Carlo a five-step downstream process whose step yields are
normally distributed with means 0.90, 0.85, 0.80, 0.85, 0.90 and
standard deviation 0.03, truncated to $[0,1]$. Report the distribution
of total recovery.
\end{exercise}

\paragraph{Discussion.}
Mean total recovery is approximately the product of the means:
$0.90 \times 0.85 \times 0.80 \times 0.85 \times 0.90 \approx 0.468$.
Standard deviation, to first order from error propagation,
$\sigma_{\text{tot}}/\mu_{\text{tot}} \approx \sqrt{\sum (\sigma_i/\mu_i)^2}$,
with $\sigma_i/\mu_i$ in the range 0.033-0.038, gives
$\sigma_{\text{tot}}/\mu_{\text{tot}} \approx \sqrt{5 \times 0.035^2}
\approx 0.078$, so $\sigma_{\text{tot}} \approx 0.037$. The Monte
Carlo distribution is approximately Gaussian (5 multiplicative steps
is enough for the central limit theorem to deliver a near-normal
shape on this scale), with a slight left skew from the truncation
near 1.0. The 5th and 95th percentiles are roughly
$0.41$ and $0.53$. The take-away: even a modest 3-point standard
deviation on each step yield gives a $\pm 12\%$ relative spread on
total recovery, which is the planner's actual uncertainty on a
campaign budget.

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a bench-scale aerobic microbial run to produce an intracellular
enzyme. Specify reactor volume, feed composition, oxygen-transfer
target, harvest criterion, and first two downstream steps.
\end{exercise}

\paragraph{Discussion.}
A 5 L stirred-tank bench reactor (working volume 3 L). Strain:
\emph{E.\ coli} BL21(DE3) with the target gene under IPTG control.
Feed: defined minimal medium (M9 with glucose, trace elements,
thiamine) plus an exponentially fed glucose stream after a batch growth
to OD$_{600}$ 4-6. Target $k_L a \ge 300\,\text{h}^{-1}$ at
800-1000 rpm and 1.0 vvm air, with provision for oxygen enrichment
above 50\% DO. Induction at OD 25 with 0.5 mM IPTG, harvest after
4-6 h induction at OD 50-100. First downstream step: continuous
centrifugation (2 L/min flow, 12000 g) to a wet-cell paste. Second
downstream step: high-pressure homogenisation at 800 bar in two
passes, followed by centrifugation to remove debris and recover
inclusion bodies. The harvest criterion is OD plateau coupled with a
target intracellular protein band in SDS-PAGE; the operational
fall-back if the OD plateau arrives earlier than expected is a
short-induction harvest at OD 30.

\begin{exercise}
Design a mammalian-cell process for a secreted protein at concept
level. Specify batch or perfusion mode, cell-retention strategy if any,
principal quality attributes, and the scale-up variable you would
protect first.
\end{exercise}

\paragraph{Discussion.}
A CHO fed-batch in a 2000 L single-use stirred-tank reactor for a
secreted IgG monoclonal antibody. Cell-retention not used (fed-batch
not perfusion); harvest at peak cell density (15-20$\times$10$^6$
cells/mL) after 12-14 days. Principal quality attributes:
glycosylation pattern (G0F / G1F / G2F ratios, sialylation,
afucosylation), aggregation $\le 2\%$, host-cell protein
$\le 100\,\text{ppm}$, host-cell DNA $\le 10\,\text{ng/dose}$,
charge variant distribution. Scale-up variable protected first: tip
speed (because shear-induced cell lysis releases proteases and
intracellular impurities that hurt glycoform quality). Secondary
protected variable: dissolved CO$_2$, because mass-transfer limits
on CO$_2$ stripping at scale alter pH control and metabolism.
Downstream first step: depth filtration / centrifugation; capture
Protein A; viral inactivation low-pH hold; AEX flow-through for
viral clearance; CEX bind-and-elute; nanofiltration; UF/DF; final
formulation and fill. Total expected yield 50-65\%, titer 3-5 g/L
in mid-2020s state of the art \cite{paper:v6c10-li-cho-cellline-2010}.

\begin{exercise}
Design a sterile sampling procedure for a small bioreactor. Draw the
sterile boundary and identify every operation that crosses it.
\end{exercise}

\paragraph{Discussion.}
The sterile boundary includes the reactor head plate, jacket-isolated
seals on each probe, the sterile inlet filter, the sparger and gas
line downstream of its sterilising filter, the harvest valve, and the
sample port. A defensible sterile sampling procedure has three
sterile components: (1) a sample valve at the broth surface or
side-stream loop, with a steam barrier (a small steam-in-place loop
between the broth and the user) that is opened to the steam side
before and after sampling; (2) a sample bottle pre-sterilised and
double-sealed, with a septum or aseptic connector; (3) an SOP that
opens the steam barrier, flushes a small purge volume, draws the
sample, closes the valve, re-steams the barrier, and labels the bottle.
Every operation that crosses the boundary is logged: number and
volume of samples per run, time, operator, and any deviation. The
sample port is in fact crossed every time it is opened; the
engineering question is whether the crossing is procedural (designed
in) or accidental (unrecorded).

\begin{exercise}
Design a household fermentation experiment that can distinguish
temperature-limited from substrate-limited fermentation. State the two
experimental arms, the measured variables, and the expected signatures.
\end{exercise}

\paragraph{Discussion.}
Use a sugar-yeast CO$_2$ fermentation. Arm A (substrate-titration):
keep temperature constant at $25\,\degC$ in three identical jars,
vary initial sugar concentration ($50, 150, 250\,\text{g/L}$) at fixed
yeast inoculum. Arm B (temperature-titration): keep sugar at
$150\,\text{g/L}$ in three identical jars, run at $15$, $25$, and
$32\,\degC$. Measure: jar mass on a kitchen scale once per day. Plot
$dm/dt$ (the mass-loss rate, equivalent to the CO$_2$ release rate).
Expected signatures: if substrate-limited, the early rate scales
roughly with initial sugar; the run finishes when sugar runs out and
the curves all approach a common cumulative loss equal to about 0.49
times the initial sugar mass. If temperature-limited, the rates
differ at all times by an Arrhenius factor ($Q_{10} \approx 2$) and
the run finishes earlier at higher temperature. The diagnostic is the
shape of the rate curve: substrate-limited produces a fast early
phase and a long tail; temperature-limited produces a near-constant
rate that ends abruptly when substrate is gone.

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A production run reaches the expected titer but fails a late
purity specification. State three hypotheses, one upstream, one
downstream, and one analytical, and the evidence you would seek for
each.
\end{exercise}

\paragraph{Discussion.}
Upstream hypothesis: a cell-line drift event (a sub-population
producing a modified glycoform or a clipped variant). Evidence to
seek: the cell-bank passage number for this run, intracellular protein
analytics on retained samples, and a charge-variant or glycoform
comparison against historical batches. Downstream hypothesis: a
chromatography column degradation (a resin past its qualified cycle
count, a column with damaged distribution, or a buffer error). Evidence
to seek: column-cycle log, in-process chromatograms compared against
historical baselines, buffer-conductivity and pH logs. Analytical
hypothesis: a method robustness failure (a reference standard at end
of qualified life, a column from a new lot with different
selectivity, an integration parameter shift). Evidence to seek: system
suitability data, reference-standard QC trend, and re-analysis of
retained sample with an orthogonal method (mass spectrometry, gel,
size-exclusion). A defensible deviation investigation tests all three
in parallel rather than accepting the first plausible explanation.

\begin{exercise}
A scale-up team proposes constant geometric similarity from
$2\,\text{L}$ to $2000\,\text{L}$ and no other criterion. State why
this is incomplete and name the failure mode you would protect first
for a fast aerobic bacterial process.
\end{exercise}

\paragraph{Discussion.}
Constant geometric similarity fixes the vessel shape but does not
fix the operating point. At the same impeller-speed-to-diameter ratio,
$P/V$ falls as $V^{-2/3}$ and $k_L a$ falls correspondingly. A
2 L reactor at $P/V = 1000\,\text{W}/\text{m}^3$ scales to a 2000 L
reactor at $P/V = 100\,\text{W}/\text{m}^3$, and $k_L a$ drops from
about $100\,\text{h}^{-1}$ to about $40\,\text{h}^{-1}$. For a fast
aerobic bacterial process the OUR is the protected failure mode:
oxygen-transfer limit, then anaerobic excursion, then overflow
metabolism, then yield loss. The correct rule is constant $k_L a$
or, more pragmatically, a constrained constant-$P/V$ scale-up with
oxygen enrichment and pressure control as backstops. The team's
proposal therefore must add at least one transport-protecting rule
beyond geometry.

\begin{exercise}
A chemostat used for production has run stably for 30 days, but product
titer is drifting downward while biomass remains constant. State three
biological explanations and two process explanations.
\end{exercise}

\paragraph{Discussion.}
Biological explanations:
(1) selection for non-producing variants: the production phenotype
is a metabolic burden, and any spontaneous loss of the production
plasmid (or mutation in the production gene) confers a growth
advantage; this accumulates as the chemostat is in effect an
evolution experiment.
(2) loss of induction effectiveness over time: the inducer is consumed
or degrades, or the strain has down-regulated the induction-response
machinery.
(3) gradual accumulation of an inhibitor (autoinducer, toxic
metabolite, secreted protease that degrades the product) whose
concentration drifts up while biomass is fixed.

Process explanations:
(1) feed composition drift: a trace nutrient is being depleted from
the medium stock that was not refilled, or the feed pH has drifted.
(2) downstream sampling artefact: the in-line product assay has
drifted (probe fouling, reference-standard drift, calibration shift)
and the apparent titer drop is in the measurement, not the process.

The correct response is to run a side-by-side comparison with a
fresh inoculum from the master cell bank, a fresh feed lot, and an
orthogonal product assay.

\begin{exercise}
An operator finds that a batch record is incomplete for a hold-time
step, but all final analytical tests pass. Explain why release is still
an engineering decision rather than a purely analytical one.
\end{exercise}

\paragraph{Discussion.}
The principle ``the batch record is part of the product'' applies.
The release decision is whether the manufactured material has a
documented identity, history, and provenance that match the
specification. If the hold-time record is incomplete, the analytical
panel can tell us about the current state of the material but not
about whether the material was processed within its qualified hold
window. A long hold can change biological products through
proteolysis, aggregation, deamidation, or oxidation, none of which is
necessarily caught by the release-test panel (which tests against
specifications, not against every possible time-dependent change). The
defensible response is a deviation investigation that reconstructs
the hold time from secondary evidence (timestamped sample logs,
operator interviews, room-environment logs, raw-data files from
in-process probes), an extended analytical panel that addresses the
time-dependent product-quality risks specifically, and a documented
decision on release that records what evidence supported it. Release
without that reconstruction is release of a product whose history is
unknown, and the engineering question is whether the supply chain
should ever release a product with that gap, irrespective of the
particular analytical results.
