\chapter{Design studios: integrated case studies}
\label{vol11:ch12}

\epigraph{The studio is the discipline applied to a complete problem; the chapters are the discipline applied to its parts.}{}

\chapmeta{Half-life: medium-life. Archetypes: ethics. Exercise target: 10. Project track: hybrid.}

\noindent
The preceding chapters of this volume developed the design discipline element by element: requirements, user research, conceptual design, detailed design, simulation, prototyping, testing, reviews, lifecycle, ethics. The studios of this chapter integrate the elements: each studio is a complete design problem worked at depth, with all elements engaged simultaneously rather than sequentially. The studios are extended; this chapter sketches the problems, the design considerations, the integrated approach, and the lessons each surfaces. The reader who completes one studio end-to-end has integrated the volume's discipline into working practice.

\section{Studio 1: a pump for a low-resource environment}\pathtag{core}

The problem: design a water pump for use in a community without reliable grid electricity, without industrial-quality replacement parts, with users who have limited engineering training, and with a need for delivery of $\sim 2{,}000$ litres of clean water per day to serve roughly 300 people.

The studio integrates:

\textbf{Requirements (Chapter 2).} The functional requirements: deliver $\sim 2{,}000$ litres per day; operate without grid electricity; operate in the local water-source conditions (depth; sediment; chemistry). The non-functional requirements: maintainable by local users with locally-available tools; manufacturable by local or accessible manufacturing capability; affordable within the community's resources; reliable over multi-year operation without specialist service.

\textbf{User research (Chapter 3).} The user population: the community members who will operate the pump; the maintainers; the children and adults who will draw water. The contexts: the social structure around water access; the existing water-management practices; the failure modes of prior interventions. The discipline includes engagement with the community as the design's stakeholders, not as recipients.

\textbf{Conceptual design (Chapter 4).} Multiple concepts generated: hand pumps (rope; piston; diaphragm); foot pumps; pedal-powered pumps; solar-powered pumps; wind-powered pumps; water-ram pumps using flowing-water energy; combinations. Concept selection based on the requirements: the appropriate concept depends on the water-source characteristics, the community's preferences, and the lifecycle considerations.

\textbf{Detailed design (Chapter 5).} The selected concept developed in detail. For a hand-pump example: cylinder dimensions, piston design, valve design, materials selection (corrosion resistance in the local water; availability locally; UV resistance for any exposed components); seals and gaskets (specifications that can be sourced or substituted locally); fasteners (standard sizes locally available); pumping mechanism (mechanical advantage suitable for the users' strength).

\textbf{Simulation (Chapter 6).} Performance modelling: discharge rate as a function of pumping rate; pressure capability at various lifts; energy requirements per litre. Lifecycle modelling: wear rates of the principal components; expected service life; maintenance interval predictions.

\textbf{Prototyping (Chapter 7).} Prototype iterations: lab prototype to verify principle; field prototype installed at a test community for evaluation; refined prototype based on field-learning; multi-unit pilot to validate manufacturing process.

\textbf{Testing (Chapter 8).} Performance testing under representative conditions; durability testing at accelerated wear rates; failure-mode testing under abuse; field testing at communities with structured data collection.

\textbf{Review and certification (Chapter 9).} Design reviews at each stage; certification against any applicable standards (drinking-water standards; structural standards); engagement with the community as the substantive validation body.

\textbf{Manufacturability and sustainability (Chapter 10).} Manufacturing process compatible with local or accessible manufacturing capability; maintainability by local users; end-of-life pathway (the pump's eventual replacement; the materials' eventual disposition); environmental footprint over lifecycle.

\textbf{Ethics (Chapter 11).} The community's engagement as partners rather than as recipients; the design's accountability to the community's needs rather than to the funder's preferences; the long-term sustainability of the intervention versus the short-term performance.

The studio's expected outcome: a complete design pack for a low-resource-environment water pump, with the analytical work, the prototype evidence, and the community-engagement evidence to support the design's deployment. The pack would be reviewable by a competent design panel and would be deployable through a community-engagement programme.

The cases this studio engages with: the substantial humanitarian-engineering literature on water-pump design (the Engineers Without Borders projects; the IDEAS academic programmes; the Practical Action publications; the academic literature on appropriate technology); the specific failure modes documented in the field (pumps that broke and were not repaired; pumps that worked but did not serve actual community patterns; pumps that produced contamination; pumps that excluded user groups). The studio's discipline is to draw on the literature and to address its lessons rather than to repeat its failures.

\section{Studio 2: an autonomous mobile robot for a defined environment}\pathtag{standard}

The problem: design an autonomous mobile robot for a defined environment such as a warehouse, a hospital corridor, a factory floor, or a research laboratory. The robot must perform a defined task (material transport; cleaning; inspection; safety monitoring) safely in the presence of human workers, in conditions the design specifies, with the autonomy level the design specifies.

The studio integrates:

\textbf{Requirements (Chapter 2).} Functional: the task the robot performs; the speed; the payload; the operating envelope. Non-functional: safety (interaction with humans; collision avoidance; fail-safe behaviour); reliability (uptime; mean time between failures); usability (interface for setup and supervision; reaction to anomalies); regulatory (compliance with applicable robot safety standards; industrial safety standards; data-protection if the robot collects information).

\textbf{User research (Chapter 3).} The user population: the workers in the environment whose work the robot will affect; the supervisors who will deploy and monitor the robot; the maintenance personnel. The contexts: the physical environment (layout; obstacles; lighting; surfaces); the social dynamics (acceptance of the robot; trust; coordination with human work); the workflow patterns the robot must integrate with.

\textbf{Conceptual design (Chapter 4).} Concepts varying in: drive mechanism (wheeled; tracked; legged); sensing modalities (lidar; vision; ultrasonic; tactile); autonomy approach (full autonomy; supervised autonomy; teleoperation with autonomy support); the human-interaction model (autonomous with humans avoiding the robot; collaborative with humans and robot sharing space; teleoperated with humans guiding the robot).

\textbf{Detailed design (Chapter 5).} The mechanical platform; the sensing hardware; the computing platform; the software architecture; the user interface; the safety architecture. The detailed design integrates Volume VIII (machines), Volume IX (systems), and Volume VII (information).

\textbf{Simulation (Chapter 6).} Software-in-the-loop simulation of the robot's behaviour in the target environment; hardware-in-the-loop simulation for safety-critical functions; full-environment simulation for system-level evaluation. The simulation is particularly important for an autonomous system whose physical testing under all conditions is impractical.

\textbf{Prototyping (Chapter 7).} Software prototypes for algorithms; hardware prototypes for mechanical platform; integrated prototypes for end-to-end function; field-deployed prototypes for actual-environment validation.

\textbf{Testing (Chapter 8).} Functional testing across the operating envelope; safety testing across hazardous scenarios; performance testing for the specified tasks; reliability testing through extended operation; usability testing with representative supervisors.

\textbf{Review and certification (Chapter 9).} Reviews at each stage; certification against ISO 13482 (personal-care robots), ISO 10218 (industrial robots), or the applicable standards; safety case documentation; risk assessment per ISO 31000.

\textbf{Manufacturability and lifecycle (Chapter 10).} Manufacturing process for the robot platform; software-deployment and -update infrastructure; maintenance and service approach; end-of-life pathway including disposition of any collected data.

\textbf{Ethics (Chapter 11).} The robot's effect on the human workers (job displacement; work-rate pressure; surveillance side effects); the data the robot collects and how it is used; the consequences of robot failures for human workers; the dual-use considerations if the robot's design could be adapted to surveillance or military applications.

The studio's expected outcome: a complete design pack for the autonomous mobile robot, including the safety case that supports its deployment in the defined environment. The cases this studio engages with: the substantial literature on industrial robotics safety (the Robotics Industries Association standards; the IEC standards); the autonomous-vehicle safety literature (which addresses analogous problems at larger scale); the specific accident records (industrial robot fatalities; autonomous-vehicle incidents; service-robot failures).

\section{Studio 3: a small renewable energy installation}\pathtag{standard}

The problem: design a renewable energy installation (solar PV; small wind; micro-hydro; biomass) for a specified scale (single-house; community; small-industry) with consideration for the grid context (off-grid; grid-tied; hybrid), the storage requirements, the load characteristics, the regulatory framework, and the lifecycle.

The studio integrates:

\textbf{Requirements (Chapter 2).} Functional: the energy delivery target (peak and average power; daily and seasonal patterns); the reliability target (acceptable outage frequency and duration); the grid interface if any. Non-functional: cost (per kWh delivered); footprint (land area; visual; noise); environmental (lifecycle emissions; end-of-life materials); maintainability; resilience to extreme weather and other shocks.

\textbf{User research (Chapter 3).} The user population: the household or community that uses the energy; the operators if any (off-grid systems often require operator engagement); the maintainers. The contexts: the local climate (solar resource; wind resource; water resource as applicable); the existing energy infrastructure; the regulatory environment; the local capacity for installation and maintenance.

\textbf{Conceptual design (Chapter 4).} The energy-source mix (solar; wind; hydro; biomass; hybrid combinations); the storage approach (lithium-ion battery; lead-acid; pumped hydro for larger scales; thermal storage); the load-management approach (demand response; load-shedding; time-of-use shifting); the grid interface if any (net-metering; grid-forming; islanding).

\textbf{Detailed design (Chapter 5).} For a solar PV example: PV array sizing and configuration; inverter selection; battery sizing; balance-of-system components; mounting structures; electrical distribution; monitoring; controls. The detailed design integrates Volume IV (energy) and Volume VIII (power electronics, Vol VIII Ch 9).

\textbf{Simulation (Chapter 6).} Energy production modelling using solar-radiation databases and weather data; load modelling; storage state-of-charge modelling; full-system simulation for performance prediction over multi-year operation. The PVsyst, HOMER Pro, or equivalent software supports this work.

\textbf{Prototyping (Chapter 7).} The prototyping is partly modelling-based (the full system is simulated rather than built at scale); a pilot installation at a representative site provides field-validation data; the prototyping cycle is longer than for products because the validation depends on multi-month operation.

\textbf{Testing (Chapter 8).} Component testing to specifications; system commissioning testing; performance verification over time. The testing extends through the installation's early operational period.

\textbf{Review and certification (Chapter 9).} Reviews of the engineering design; certification of the components (UL standards; IEC standards; the applicable national equivalents); permitting and inspection per the local jurisdiction.

\textbf{Manufacturability and lifecycle (Chapter 10).} The installation's manufacturing is typically assembly of standard components rather than custom manufacturing; the maintainability over 20-30 year service life; the end-of-life pathway for the components (PV-module recycling; battery recycling; inverter disposition); the lifecycle environmental footprint compared to the displaced fossil-fuel generation.

\textbf{Ethics (Chapter 11).} The installation's contribution to the broader energy transition (Volume X Chapter 15); the equity considerations (who benefits from the installation; who is excluded); the long-term commitment to the operator's continuation; the dependence on the supply chain for replacement components.

The studio's expected outcome: a complete design pack for the renewable energy installation, suitable for engineering review and for the permitting and procurement process. The cases this studio engages with: the substantial literature on renewable-energy system design; the failure-mode documentation from prior installations (premature inverter failures; underperforming arrays; battery degradation; wildfire risk; storm damage); the regulatory environment for grid interconnection and energy markets.

\section{Studio 4: a software-physical system with safety implications}\pathtag{core}

The problem: design a software-physical system whose failures have safety implications: a controlled medical device; a building automation system; an industrial process controller; a vehicle subsystem; a critical infrastructure controller. The system must integrate software and physical engineering, must address safety as a primary requirement, and must satisfy regulatory and certification requirements appropriate to its safety class.

The studio integrates:

\textbf{Requirements (Chapter 2).} Functional: the controlled function (the physical process the software controls; the operator interactions; the data exchanges). Non-functional: safety (the failure modes; the safety integrity level; the required mitigations); security (the threats; the protections; the authentication and authorisation); reliability (availability; mean time between failures); regulatory (the applicable standards: IEC 61508 for industrial; ISO 26262 for automotive; DO-178C for avionics; IEC 62304 for medical; etc.).

\textbf{User research (Chapter 3).} The user population: operators; maintenance staff; emergency responders; patients (for medical devices). The contexts: the operational environment; the failure conditions and responses; the integration with other systems and procedures.

\textbf{Conceptual design (Chapter 4).} The system architecture: the safety-critical functions versus the non-safety-critical; the redundancy approach; the partitioning between software and hardware safety mitigations; the diversity (or lack of it) in the safety architecture; the human-in-the-loop approach.

\textbf{Detailed design (Chapter 5).} The control algorithms; the software architecture; the hardware platform; the safety case documentation; the failure-mode-and-effects analysis; the human-machine interface; the security architecture.

\textbf{Simulation (Chapter 6).} Software simulation of the control algorithms; hardware-in-the-loop simulation for safety verification; full-system simulation for operational evaluation. The simulation is particularly important for safety-critical systems where physical fault injection is impractical.

\textbf{Prototyping (Chapter 7).} Software prototypes for algorithm development; integrated prototypes for end-to-end function; safety-relevant prototypes for failure-mode validation; pre-clinical (for medical) or pre-deployment (for industrial) evaluation.

\textbf{Testing (Chapter 8).} Software testing per the applicable standard (DO-178C requires MC/DC coverage for highest assurance level; IEC 62304 requires lower coverage but extensive documentation); integration testing; system testing; safety-case verification testing; security testing including penetration testing.

\textbf{Review and certification (Chapter 9).} Reviews at each stage with safety experts and certification engineers; safety case submission to the appropriate authority; iterative engagement with the regulator; sign-off by the responsible safety engineer.

\textbf{Manufacturability and lifecycle (Chapter 10).} The system's manufacturing (hardware production; software image deployment); the maintainability over service life; the software-update infrastructure; the end-of-life pathway; the long-term support commitments.

\textbf{Ethics (Chapter 11).} The risk acceptance for affected users; the disclosure of residual risks; the data the system collects and its use; the dual-use considerations for software the system uses; the engineer's individual sign-off as the substantive commitment to the design's safety.

The studio's expected outcome: a complete design pack for the safety-critical software-physical system, including the safety case documentation that supports its certification submission. The cases this studio engages with directly include the Therac-25 case (Volume X Chapter 5), the 737 MAX MCAS case (Volume X Chapter 8), the Toyota unintended-acceleration cases, and the broader medical-device, industrial-control, and automotive safety-critical literature. The discipline of the studio is to internalise the lessons from these cases rather than to repeat their failures.

\section{Failure: each studio includes its own failure-mode analysis}\pathtag{standard}

Each studio includes failure-mode analysis as a substantive part of the design work, not as a checklist item. The analyses address the specific failure modes the studio's design class is prone to:

\textbf{Studio 1 (low-resource pump).} Failure modes include: mechanical wear of the principal moving parts; corrosion of metallic components; UV degradation of polymer components; user-induced damage from misuse; vandalism; contamination of the water source; the maintenance organisation's collapse; the supply chain's interruption for replacement parts. The analysis identifies the highest-consequence failure modes and the design responses; the responses inform the detailed design.

\textbf{Studio 2 (mobile robot).} Failure modes include: collision with a human worker; collision with infrastructure; tipping; loss of localisation; loss of communication; battery depletion; software errors causing incorrect motion; emergency-stop failure; perception failures (the robot misclassifies what it sees); software updates introducing regressions. The analysis informs the safety architecture, the operating-envelope restrictions, and the documentation.

\textbf{Studio 3 (renewable installation).} Failure modes include: PV module degradation; inverter failure; battery degradation; storm damage; lightning damage; theft; insurance and warranty failures; underperformance against design assumptions; islanding failures during grid outages; ground faults; fire risks. The analysis informs the design margins, the protection systems, the maintenance regime, and the insurance approach.

\textbf{Studio 4 (safety-critical software-physical).} Failure modes include the full Volume X taxonomy: hardware failure; software defects; concurrency and timing issues; integration mismatches; user-interaction errors; cyberattack; supply-chain compromises; certification gaps; in-service degradation. The analysis informs the safety case, the certification submission, and the operational monitoring.

The studios' failure-mode analysis discipline includes:

\textbf{Systematic enumeration.} Each studio's failure modes are enumerated through structured techniques (FMEA, HAZOP, fault tree, STAMP). The enumeration is comprehensive within the design's scope; the discipline addresses what could go wrong rather than what is convenient to address.

\textbf{Consequence assessment.} Each failure mode's consequences are characterised: who is affected; how severely; over what timescale; with what recovery options. The consequences inform the priority of the response.

\textbf{Likelihood estimation.} Each failure mode's likelihood is estimated; the estimation uses operational data, engineering judgment, and structured uncertainty quantification. The likelihood combines with the consequence to produce the risk assessment.

\textbf{Mitigation design.} For each significant failure mode, the design includes mitigations: prevention (the failure does not occur); detection (the failure is detected if it occurs); mitigation (the consequences are reduced when the failure occurs); recovery (the system returns to acceptable state after the failure).

\textbf{Verification of mitigations.} The mitigations are verified in the testing programme; the verification confirms the mitigations operate as designed.

\textbf{Documentation.} The failure-mode analysis is documented as part of the design pack; the documentation supports certification, service support, and forensic analysis if failures occur in service.

The discipline of integrated failure-mode analysis (rather than checklist-style failure-mode analysis) is what distinguishes substantive design from theatre. The studios are intended to develop the discipline through extended application; the alternative (failure-mode analysis as a documentation requirement) produces documents that meet certification requirements without producing safer designs.

\section{Reflection: what each studio teaches that the chapters could not}\pathtag{core}

The chapters of this volume developed the design discipline element by element. The studios integrate the elements; the integration teaches what element-by-element treatment cannot:

\textbf{The trade-offs that exist only at integration.} Each design element has its own discipline; the trade-offs between elements emerge only when the elements are integrated. Tolerance allocation (Chapter 5) trades off against manufacturability (Chapter 10); reliability requirements (Chapter 2) trade off against cost; user-research findings (Chapter 3) trade off against schedule. The trade-offs are made by the design team; the studios provide the working practice in making them.

\textbf{The iteration that exists only across elements.} The design's iteration is not within elements but across them: requirements change in response to user research; concepts change in response to detailed analysis; detailed designs change in response to test findings. The studios provide the practice in the cross-element iteration.

\textbf{The substantive ambiguity of real problems.} The chapters present examples that are clean enough to teach a specific point; real problems are ambiguous, ill-defined, and require the engineer to bound the problem before solving it. The studios provide the practice in bounding real problems.

\textbf{The cumulative effect of small decisions.} Each chapter develops a discipline that addresses specific design decisions; the cumulative effect of many small decisions across the chapters is not visible at the chapter level. The studios make the cumulative effect visible; designs that look good in element-by-element review can be substantively inadequate at integration.

\textbf{The engineer's developing judgment.} The chapters develop techniques; the studios develop judgment about when to apply which techniques, how much depth to invest in each, when to commit and when to iterate. Judgment develops through extended practice; the studios provide the extended practice.

\textbf{The engagement with stakeholders.} The chapters describe stakeholder engagement abstractly; the studios require it concretely. The engineer who has navigated the conflicting requirements of multiple stakeholders, the resistance of cost-conscious project managers, the difficulty of communicating across disciplines has integrated the discipline more substantively than chapter-by-chapter reading can.

\textbf{The professional commitment.} The chapters describe the engineering profession's commitments; the studios produce design packs to which the engineer's name would be attached. The professional commitment that the signature represents (Chapter 9) is exercised in the studios at substantive scale.

The studios are extended efforts: each is months of substantive work. The volume's expectation is not that the reader complete all four; the reader who completes one studio end-to-end has internalised the volume's discipline in a way that less-extended engagement cannot achieve. The capstone (Chapter 13) is the further extension: the reader's choice of an engineering problem of meaningful scale, worked at the integrated discipline the studios prefigure.

\begin{archetype}[ethics]
The studios are the volume's expression of engineering's integrated discipline. The chapters teach the elements; the studios teach the integration; the integration is the engineering profession's substantive contribution. The discipline that the studios develop is the discipline that distinguishes the engineer from the technician: the substantive engagement with complete problems, under realistic constraints, with professional accountability for the outcomes.
\end{archetype}

\begin{principle}[Integration is the substantive engineering]
The chapters of this volume teach the discipline element by element; the studios integrate the elements; the integration is where the substantive engineering happens. Designs that look adequate at element review can be inadequate at integration; designs that integrate well are designs whose elements were developed with integration in mind from the start. The discipline of integration is the discipline of the studios and, beyond them, of the capstone.
\end{principle}

\begin{project}[Complete one studio end-to-end]
Select one of the four studios. Work it end-to-end over an extended period (suggested: 3 months of substantive part-time effort, or 1 month of full-time). The deliverable is the complete design pack: requirements; user-research findings; concepts and selection; detailed design; simulation evidence; prototype evidence; test results; review records; lifecycle considerations; failure-mode analysis; ethical considerations; ethics statement specific to the studio. The pack is reviewable by a competent design panel.

\textbf{Hazard.} The studio is substantial; estimate before starting how much the reader can commit. The discipline of completing one studio fully is more valuable than the partial completion of several. The substantive learning is in the integration, not in the breadth.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Studio 1 specifies $\sim 2{,}000$ litres of clean water per day for $\sim 300$ people. Compute the per-person daily allocation and compare to WHO minimum (50 litres per person per day for basic survival) and to WHO recommended (100 litres per person per day for sustainable health). Identify what the studio's allocation supports.
\end{exercise}

\begin{exercise}
Studio 3 (renewable installation) for a typical European household uses 3500 kWh per year. With Northern European average insolation of 1000 kWh/m$^2$ per year and PV module efficiency of 20\%, compute the PV array area required to meet the annual energy demand. Identify the considerations the simple calculation does not include.
\end{exercise}

\begin{exercise}
Studio 4 (safety-critical software-physical system) requires DO-178C Level A software with 100\% statement coverage and 100\% MC/DC coverage. Estimate the test-development effort required, given the engineering-industry rule of thumb of approximately 5 person-hours per line of safety-critical code through coverage and review.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the total engineering effort to complete Studio 1 end-to-end (the design pack only, not the deployed pump). Identify the dominant effort categories.
\end{exercise}

\begin{exercise}
Estimate the certification effort for Studio 4 in a medical-device context (Class II 510(k) submission or Class III PMA submission). Compare to the design-development effort.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
For Studio 2 (mobile robot), design the safety architecture: the safety-rated functions; the redundancy; the fail-safe behaviours; the operator-override mechanisms; the safety-case documentation. Identify the standards (ISO 13482 or ISO 10218 or analogous) that apply and the evidence the standards require.
\end{exercise}

\begin{exercise}
For Studio 3 (renewable installation), design the failure-tolerance strategy: how the system handles inverter failure; battery failure; PV module failure; weather extremes; grid outages. Specify the residual reliability under each failure mode.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Discuss the ethical considerations that distinguish Studio 1 (low-resource environment) from Studio 4 (safety-critical software-physical). Identify what each studio's ethical analysis must address that the other's does not, and what they share.
\end{exercise}

\begin{exercise}
Argue: the studio approach is the only way to develop integrated design judgment; chapter-by-chapter learning produces engineers who know the elements but cannot integrate them. Argue the opposite: chapter learning is the prerequisite; studio integration is supplementary practice. Identify what determines which view is correct for a given engineer's development.
\end{exercise}

\begin{exercise}
A design organisation has historically deployed Studio-1-class solutions in low-resource environments. Many of the deployments are no longer functional. Discuss what would have been required of the organisation's design discipline to produce sustained interventions rather than projects whose deliverable was the installation rather than the sustained service.
\end{exercise}
