\chapter{Proteins as machines}
\label{vol06:ch04}

\epigraph{The native conformation is determined by the totality of
interatomic interactions and hence by the amino acid sequence, in a
given environment.}{Christian B. Anfinsen, Nobel lecture (1972), in
\emph{Science} 181:223--230 \cite[p.~223]{paper:anfinsen1973-protein-folding}.}

\chapmeta{Half-life: foundational. Archetypes: control, scaling. Exercise target: 25. Project track: visualisation and structure-function analysis of three proteins from the Protein Data Bank using free molecular-viewer software. Hazard class: none (computational work only). Reader path default: standard, with sections explicitly tagged. Named cases: none required; misfolding diseases (Alzheimer's, prion diseases, sickle cell) are described as mechanism classes rather than named registry events.}

\noindent
Proteins are the cell's machines. Each protein is a polypeptide chain
that has folded into a specific three-dimensional shape, and the
shape is what makes the protein do what it does. The chapter develops
the protein-as-machine reading of cellular biology: it walks from the
amino-acid primary structure through the folding process to the
functional classes (enzymes, motors, channels, pumps), introduces
allosteric regulation as the cell's principal mechanism for tuning
enzymatic activity, and closes with the failure modes that arise when
proteins fold incorrectly or fail to maintain their structure under
load.

\section{Amino acids and primary structure}\pathtag{core}

A \engterm{protein} is a linear polymer of amino acids joined by
peptide bonds. The cell uses $20$ standard amino acids, plus the two
non-standard amino acids selenocysteine and pyrrolysine that some
organisms encode through specific recoding mechanisms. Each amino
acid has the same backbone (an amino group, a central $\alpha$-carbon,
and a carboxylic acid group) and a distinct side chain. The $20$
side chains span a chemical range that includes hydrophobic
(alanine, valine, leucine, isoleucine, methionine, phenylalanine,
tryptophan), polar uncharged (serine, threonine, asparagine,
glutamine, tyrosine), charged (aspartate and glutamate negative,
lysine and arginine positive, histidine positive at physiological pH
of $7.4$ with $\text{pK}_a \approx 6.0$), special (glycine with no
side chain, proline with a cyclic side chain, cysteine with a thiol
that can form disulfide bonds) \cite[ch.~3]{text:lehninger-biochemistry}.

\engterm{The peptide bond}. The peptide bond is an amide bond between
the carboxyl carbon of one amino acid and the amino nitrogen of the
next. The bond has partial double-bond character (delocalised
resonance with the adjacent carbonyl), which makes it planar and
restricts the polypeptide backbone to two rotational degrees of
freedom per residue: the $\phi$ angle (rotation around the
$\text{N}-\text{C}_\alpha$ bond) and the $\psi$ angle (rotation around
the $\text{C}_\alpha-\text{C}$ bond). The allowed combinations of
$\phi$ and $\psi$ form the \engterm{Ramachandran plot}, with three
principal regions corresponding to right-handed $\alpha$-helix
($\phi \approx -60^{\circ}$, $\psi \approx -45^{\circ}$),
$\beta$-sheet ($\phi \approx -135^{\circ}$, $\psi \approx +135^{\circ}$),
and left-handed $\alpha$-helix (rare, found in glycine-rich loops).
The plot's empty regions correspond to side-chain steric clashes
that no amino acid can fit. The Ramachandran plot is the working
diagnostic for whether a proposed protein structure is sterically
plausible.

\engterm{Primary, secondary, tertiary, quaternary structure}. Protein
structure is described at four hierarchical levels. \engterm{Primary
structure} is the amino-acid sequence. \engterm{Secondary structure}
is the local arrangement of the backbone into regular elements: the
$\alpha$-helix (a right-handed coil with $3.6$ residues per turn,
$0.54\,\text{nm}$ rise per turn, hydrogen bonds between the carbonyl
of residue $i$ and the amide of residue $i+4$), and the
$\beta$-sheet (parallel or antiparallel strands hydrogen-bonded
side-by-side, $0.7\,\text{nm}$ per residue along the strand). Both
secondary-structure elements were predicted by Linus Pauling and
Robert Corey in 1951 from model-building constraints, before any
high-resolution protein structure was available; the predictions were
confirmed by the first X-ray crystal structures of myoglobin and
haemoglobin in 1958 to 1960. \engterm{Tertiary structure} is the
overall three-dimensional fold of the polypeptide chain.
\engterm{Quaternary structure} is the assembly of multiple
polypeptide chains into a functional complex (haemoglobin, two
$\alpha$ and two $\beta$ chains; the proteasome, $28$ subunits; the
ribosome, $\sim 80$ proteins plus several RNAs).

\engterm{Bond budget}. The free energy of a folded protein is the sum
of contributions from: hydrogen bonds within the backbone (typically
$10$ to $30\,\text{kJ}/\text{mol}$ per hydrogen bond, although
buried hydrogen bonds in the protein interior can be stronger);
hydrophobic burial of non-polar side chains (the dominant driving
force for folding, accounting for $50$ to $100\,\text{kJ}/\text{mol}$
of stability for a typical small protein); electrostatic interactions
between charged side chains ($5$ to $20\,\text{kJ}/\text{mol}$ for
buried salt bridges); van der Waals packing in the protein core; and
disulfide bonds in extracellular proteins ($\sim 100\,\text{kJ}/\text{mol}$
each, but only present in oxidising environments). The net stability
of a typical folded protein over its unfolded reference state is
$20$ to $60\,\text{kJ}/\text{mol}$, a small number that is the
difference of two large contributions (favourable enthalpic and
hydrophobic interactions, unfavourable entropy of folding). The
small net stability is the engineering reading of protein
\engterm{marginal stability}: most proteins are just stable enough
to maintain their fold under physiological conditions, and modest
perturbations (temperature, pH, denaturants) can unfold them
\cite[ch.~4]{text:branden-tooze-proteins}.

\section{Protein folding}\pathtag{standard}

A protein's three-dimensional structure is, in most cases,
determined by its amino-acid sequence under the cell's solvent and
temperature conditions. The thermodynamic hypothesis, due to Christian
Anfinsen's 1962 ribonuclease refolding experiments, states that the
native conformation is the global free-energy minimum of the
polypeptide in its physiological environment
\cite{paper:anfinsen1973-protein-folding}. The hypothesis has been
confirmed for most small, single-domain proteins by their ability to
refold spontaneously after chemical denaturation; for larger or more
complex proteins, the folding is kinetically assisted by
\engterm{chaperone} proteins that prevent off-pathway aggregation but
do not alter the final structure.

\engterm{Levinthal's paradox and the folding funnel}. A protein of
$N$ residues with three rotational states per residue would have
$3^N$ possible conformations; for $N = 100$, the number is $5 \times
10^{47}$. If folding proceeded by random search through all
conformations, the time to find the native state would exceed the
age of the universe. Proteins, however, fold in microseconds to
seconds. The resolution, due to Wolynes, Bryngelson, and others in
the 1980s and 1990s, is that the folding-energy landscape is not
flat but has a \engterm{folding funnel} shape: the native state is
at the bottom of a broad, downhill-biased energy landscape that
channels the polypeptide toward the native state regardless of the
starting conformation. The funnel shape is itself a result of
evolutionary selection: protein sequences that fold reliably are the
ones that survived. The folding-funnel framing is the modern
engineering reading of protein folding and the basis of most
computational structure-prediction methods
\cite[ch.~6]{text:dill-bromberg-molecular-driving}.

\engterm{Folding intermediates and pathways}. Most proteins fold
through one or more intermediate states. The molten-globule
intermediate has approximately the correct secondary structure but
loose tertiary packing; subsequent steps refine the side-chain
contacts to give the native state. The classical phi-value analysis
of Fersht and others identifies the residues whose stabilising
interactions form early in the folding pathway (the folding nucleus)
versus those that form late. For many small proteins, the folding
nucleus comprises $5$ to $10$ residues whose hydrophobic core forms
in the first few microseconds and templates the rest of the fold.

\engterm{Chaperones}. The cell's protein-folding machinery includes
families of chaperone proteins that assist folding without becoming
part of the final structure. The Hsp70 family binds nascent
polypeptide chains as they emerge from the ribosome and prevents
premature aggregation; the GroEL/GroES chaperonin system in bacteria
(and the eukaryotic TRiC/CCT system) provides an isolated folding
chamber in which a misfolded protein can attempt folding without
interference from other proteins. The cell's chaperone capacity is
upregulated under heat stress (the heat-shock response), under
oxidative stress, and in cells with high protein-synthesis demand.
Chaperone failure is implicated in many of the misfolding diseases
named in the failure section.

\engterm{Computational structure prediction}. Predicting protein
structure from sequence was a fifty-year challenge that approached
practical resolution in 2020 with DeepMind's AlphaFold 2. The system
achieved approximately experimental-quality accuracy on the CASP14
benchmark (median backbone RMSD of $1\,\si{\angstrom}$ against
crystal-structure references) and has subsequently produced
predicted structures for $\sim 200\,$million proteins in the AlphaFold
database \cite{paper:jumper-alphafold2-2021}. The success has
transformed structural biology by making predicted structures
routinely available for proteins that have never been crystallised.
The system's limitations (it predicts a single static conformation,
does not handle most ligand binding well, and is less reliable for
intrinsically disordered regions) remain active research areas at
the time of writing (current as of 2025).

\begin{principle}[Sequence specifies structure under physiological conditions]
For most proteins, the amino-acid sequence under cellular solvent and
temperature conditions uniquely determines the three-dimensional
fold. The determination is thermodynamic (native state is the
global free-energy minimum) and engineered by selection (the
folding-energy landscape funnels toward the native state on the
timescale of the cell's protein-synthesis rate). The principle
makes possible the engineering programme of designing a desired
function and synthesising a sequence that folds to perform it.
\end{principle}

\section{Enzymes as catalysts}\pathtag{core}

An \engterm{enzyme} is a protein (or, in some cases, an RNA) that
catalyses a specific chemical reaction. Enzymes accelerate reactions
by factors of $10^6$ to $10^{17}$ over the uncatalysed rate, with
rate enhancements concentrated at the chemically difficult steps that
the cell needs to control \cite[ch.~6]{text:lehninger-biochemistry}.

\engterm{Catalytic mechanism}. The enzyme catalyses a reaction by
binding the substrate in a specific orientation (\engterm{transition-state
stabilisation}) and by providing a chemical environment that stabilises
the reaction transition state more than the substrate or product. The
binding pocket (the \engterm{active site}) is a small region of the
protein, usually $1\,\text{nm}^3$ or smaller, lined with the catalytic
amino-acid residues that participate in the reaction. The
transition-state stabilisation lowers the activation energy and so
accelerates the reaction.

\engterm{Michaelis-Menten kinetics}. The rate of an enzyme-catalysed
reaction depends on the substrate concentration in a saturable way
described by the Michaelis-Menten equation
\[
v = \frac{V_{\max} [S]}{K_M + [S]},
\]
where $V_{\max}$ is the maximum rate (saturating substrate), $K_M$
is the Michaelis constant (the substrate concentration at which the
rate is half-maximum), and $[S]$ is the substrate concentration. The
parameter $K_M$ ranges from $10^{-7}$ to $10^{-3}\,\text{M}$ for
typical enzymes and is usually close to the physiological substrate
concentration, so that the enzyme operates in the sensitive
intermediate region of the rate curve. The maximum turnover number
$k_{\text{cat}} = V_{\max} / [E]_{\text{total}}$ ranges from $1$ to
$10^7\,\text{s}^{-1}$. The catalytic efficiency
$k_{\text{cat}}/K_M$ ranges up to $10^9\,\text{M}^{-1}\text{s}^{-1}$;
the upper limit is set by the diffusion-limited encounter rate
between enzyme and substrate, and the most efficient enzymes
(carbonic anhydrase, acetylcholinesterase, triose phosphate isomerase)
operate at or near this limit \cite[ch.~6]{text:lehninger-biochemistry}.
The Michaelis-Menten formalism was derived by Leonor Michaelis and
Maud Menten in 1913 \cite{paper:michaelis-menten1913}.

\engterm{Catalytic strategies}. The cell uses six principal
catalytic strategies: \engterm{acid-base catalysis} (proton transfer
to or from the substrate), \engterm{covalent catalysis} (formation of
a temporary covalent enzyme-substrate intermediate), \engterm{metal-ion
catalysis} (a metal cofactor stabilises charge or activates water),
\engterm{electrostatic catalysis} (the enzyme's electrostatic
environment stabilises charged transition states), \engterm{proximity
and orientation} (the enzyme binds the substrate in the geometry
required for reaction), and \engterm{strain induction} (the enzyme
distorts the substrate toward the transition-state geometry). Most
enzymes use several strategies simultaneously; the specific
combination is set by the chemistry of the catalysed reaction.

\engterm{Enzyme classification}. The Enzyme Commission (EC)
classification organises enzymes by reaction type into six main
classes: oxidoreductases (electron transfer reactions),
transferases (group transfer reactions, including kinases that
transfer phosphate), hydrolases (hydrolytic cleavage), lyases
(non-hydrolytic cleavage), isomerases (rearrangements), and ligases
(bond formation coupled to ATP or GTP hydrolysis). The
classification, dating to the International Union of Biochemistry's
1961 report, is the standard engineering vocabulary for the cell's
catalytic catalogue.

\section{Allosteric regulation}\pathtag{standard}

The cell needs to regulate enzymatic activity in response to changing
metabolic conditions. The principal mechanism is
\engterm{allosteric regulation}: a molecule binds at a site distinct
from the active site (the \engterm{allosteric site}) and changes the
enzyme's activity by altering its conformation. The mechanism allows
the activity of an enzyme to be modulated by signals that are
chemically unrelated to its substrate.

\engterm{Two canonical models}. The MWC (Monod-Wyman-Changeux) model
of 1965 treats an allosteric enzyme as existing in two
conformational states (R, ``relaxed'' or active; T, ``tense'' or
inactive) that interconvert reversibly, with substrate binding
shifting the equilibrium toward R. The KNF (Koshland-N\'emethy-Filmer)
model treats the conformational change as occurring sequentially on
each subunit upon substrate binding (induced fit). The two models
give different specific predictions for the cooperativity curves,
and many real enzymes are intermediate between them.

\engterm{Cooperativity and the Hill equation}. Allosteric enzymes
with multiple substrate-binding sites display cooperative binding:
the binding of one substrate molecule alters the affinity for the
next. Positive cooperativity (later sites bind more tightly) gives
the steep sigmoidal binding curve characteristic of haemoglobin
oxygen binding. The empirical Hill equation
\[
\theta = \frac{[S]^n}{K^n + [S]^n}
\]
parameterises the cooperativity by the Hill coefficient $n$, which
for non-cooperative binding equals 1, for haemoglobin is approximately
$2.8$, and for the most cooperative natural systems approaches the
number of binding sites $n_{\max}$. The Hill equation is empirical;
the MWC and KNF models give it mechanistic interpretation.

\engterm{Feedback inhibition}. The cell uses allosteric regulation to
build feedback loops at the enzyme level. The end product of a
biosynthetic pathway often binds allosterically to the pathway's
first enzyme and inhibits it; when the end product accumulates, the
pathway shuts down. The bacterial aspartate transcarbamylase (ATCase)
in pyrimidine biosynthesis is the textbook example: the end product
CTP binds at a site distant from the active site and reduces the
enzyme's substrate affinity. The feedback loop is one of the cell's
fastest regulatory mechanisms (response time of seconds to minutes,
compared to hours for transcriptional regulation) and is the
mechanism of choice for stabilising biosynthetic-pathway fluxes
against perturbations.

\engterm{Covalent modification as a control variable}. Many enzymes
are regulated by reversible covalent modification of a specific
amino acid residue. Phosphorylation of serine, threonine, or
tyrosine residues by protein kinases is the most common modification;
the modification is reversed by phosphatases. Phosphorylation
changes the enzyme's activity, its subcellular localisation, or its
binding partners. The kinase-phosphatase network of a typical
mammalian cell contains $\sim 500$ kinases and $\sim 200$
phosphatases and constitutes a major fraction of the cell's
regulatory bandwidth.

\section{Motor proteins}\pathtag{standard}

The cell uses ATP-hydrolysing motor proteins for directed motion at
the molecular scale. The motor proteins are organised into three
principal families that walk along cytoskeletal tracks:
\engterm{myosin} on actin filaments, \engterm{kinesin} on microtubules
toward the plus end, and \engterm{dynein} on microtubules toward the
minus end. A fourth class, the \engterm{rotary motors} (the
mitochondrial ATP synthase, the bacterial flagellar motor), perform
rotational rather than linear motion. The engineering of motor
proteins is treated in detail by Howard
\cite[ch.~5--10]{text:howard-mechanics-motor-proteins}.

\engterm{Myosin}. The myosin family includes the muscle myosin II
that powers skeletal muscle contraction, the unconventional myosin I
that pulls membrane vesicles along actin filaments, and the myosin V
that walks processively along actin carrying cargo. Each myosin
molecule has a globular head domain that binds actin and hydrolyses
ATP, a neck region (the lever arm), and a tail that determines the
specific function (filament assembly in muscle myosin, cargo binding
in transport myosins). The mechanical work per ATP hydrolysed is
approximately $5\,\text{pN} \times 5\,\text{nm} = 25\,\text{pN} \cdot
\text{nm} = 6\,k_B T$ at $310\,\si{\kelvin}$, or about $25\,\%$ of the
$30\,\text{pN} \cdot \text{nm}$ of free energy available per ATP
under physiological conditions. The mechanical efficiency is
comparable to a well-engineered macroscopic engine.

\engterm{Muscle contraction}. Skeletal muscle is a particularly
well-characterised example. The contractile unit is the
\engterm{sarcomere}, a $2\,\si{\micro\meter}$-long structure
consisting of interdigitated thin (actin) and thick (myosin)
filaments. The myosin heads project from the thick filaments and bind
the thin filaments; ATP-driven conformational changes in the heads
slide the thin filaments relative to the thick filaments, shortening
the sarcomere. The sliding-filament model (Andrew Huxley, Hugh
Huxley, Jean Hanson, 1954) is the canonical mechanism. A single muscle
cell contains thousands of sarcomeres in series within each
myofibril and thousands of myofibrils in parallel; the macroscopic
force is the integral over the parallel myofibrils, and the
macroscopic shortening is the integral over the series sarcomeres.
The maximum force per cross-sectional area of skeletal muscle is
approximately $0.3\,\text{MPa}$, set by the density of myosin heads
and the per-head force.

\engterm{Kinesin and dynein}. The kinesin family transports cargo
along microtubules. The dimeric kinesin-1 takes $8\,\text{nm}$ steps
along the microtubule lattice, hydrolysing one ATP per step and
producing forces up to $7\,\text{pN}$. The processivity (the
expected number of steps before detachment) is $\sim 100$ for
kinesin-1, giving a typical run length of $1\,\si{\micro\meter}$.
The motor's hand-over-hand stepping mechanism, in which the two
heads alternate between bound and detached states, gives kinesin
the highly processive behaviour required for long-distance cargo
transport in neurons. Dynein operates similarly but in the opposite
direction along the microtubule.

\engterm{Rotary motors: ATP synthase}. The mitochondrial ATP synthase
(F$_1$F$_0$ ATPase) is a rotary molecular machine that synthesises
ATP from ADP and inorganic phosphate using the proton-motive force
across the mitochondrial inner membrane. The membrane-embedded F$_0$
sector contains a ring of $8$ to $15$ c-subunits that rotate as
protons pass through; the soluble F$_1$ sector contains three
catalytic $\beta$-subunits that synthesise ATP in turn as the
central stalk rotates. The motor was visualised in single-molecule
experiments by Hiroyuki Noji and colleagues in 1997 by attaching an
actin filament to the central stalk and watching it rotate. The
rotation rate is $\sim 100\,\text{revolutions}/\text{s}$ at full load
in mitochondria; the mechanical work per revolution is three ATP
synthesised, or approximately $200\,\text{pN}\cdot\text{nm}$ of free
energy stored in chemical form.

\engterm{Rotary motors: bacterial flagellum}. The bacterial flagellar
motor rotates at $100$ to $300\,\text{Hz}$ and propels the cell at
$20$ to $50\,\si{\micro\meter}/\text{s}$. The motor uses the proton
gradient across the cell membrane as its energy source (in
\emph{E.~coli}; in marine bacteria it uses sodium gradient).
Approximately $1000$ protons pass through the motor per revolution.
The motor switches direction in response to chemotactic signals from
the cell, producing the run-and-tumble swimming pattern of
\emph{E.~coli} chemotaxis (Chapter~\ref{vol06:ch01}).

\section{Membrane proteins, channels, and pumps}\pathtag{standard}

The plasma membrane (Chapter~\ref{vol06:ch01}) is a lipid bilayer
into which proteins are embedded that perform the membrane's
transport, signalling, and adhesion functions. The principal classes
of membrane protein are channels (passive conductors of ions or
small molecules), pumps (active transporters powered by ATP or a
gradient), receptors (signal transducers), and transporters
(carriers that translocate substrates by conformational change).

\engterm{Ion channels}. Channels conduct ions down their
electrochemical gradient at rates of $10^6$ to $10^8\,\text{ions}/
\text{second}$. The selectivity of a channel (whether it conducts
sodium, potassium, calcium, chloride, or some combination) is set by
a narrow region of the channel pore lined with chemical groups that
discriminate among the candidates by size and charge. The classic
example is the potassium channel selectivity filter, whose backbone
carbonyls mimic the hydration shell of a potassium ion at the cost
that a sodium ion (smaller, with stronger hydration) cannot fit
without paying a large desolvation energy. The selectivity is
typically $1000:1$ or better; the engineering of the filter is the
canonical example of size-and-coordination-based selectivity in
biology \cite[ch.~14]{text:hille-ion-channels}.

Channels are gated by various mechanisms: voltage-gated channels
respond to membrane potential changes (the sodium and potassium
channels of the action potential, Chapter~\ref{vol06:ch07});
ligand-gated channels respond to bound chemical signals (the
nicotinic acetylcholine receptor at neuromuscular junctions);
mechano-gated channels respond to membrane stretch (the touch
receptors); temperature-gated channels respond to thermal stimuli
(the TRP channels involved in pain and temperature sensation). The
gating mechanism determines the channel's role in cellular
signalling.

\engterm{Pumps}. Active transporters use an energy input (ATP or a
gradient) to move substrates against their electrochemical gradient.
The sodium-potassium ATPase (Chapter~\ref{vol06:ch01}) is the
canonical example: it exchanges $3\,\text{Na}^+$ outward for
$2\,\text{K}^+$ inward per ATP hydrolysed, against both ions'
gradients. The Ca$^{2+}$-ATPase pumps $2\,\text{Ca}^{2+}$ per ATP
out of the cytoplasm into intracellular compartments. The H$^+$-ATPases
acidify lysosomes and the plant vacuole. The cell's energy budget for
these pumps is large: the sodium-potassium ATPase alone consumes
$20\,\%$ to $30\,\%$ of the cell's resting ATP supply in many
neurons.

\engterm{Transporters}. Carriers translocate substrates by undergoing
conformational changes that alternately expose the binding site to
one side of the membrane and the other. The glucose transporter
GLUT1 is the textbook example: it binds glucose on the extracellular
side, undergoes a conformational change to expose the binding site
to the cytoplasm, releases glucose, returns to the original
conformation, and repeats. The transport rate is approximately
$1000\,\text{glucose}/\text{transporter}/\text{s}$, slower than ion
channels by three to four orders of magnitude. The slower rate is
the price the cell pays for the conformational-change mechanism
that handles polar molecules too large for simple channel passage.

\engterm{Receptors}. Membrane receptors transduce extracellular
signals into intracellular responses. The principal classes are
G-protein-coupled receptors (GPCRs, $\sim 800$ in the human genome),
receptor tyrosine kinases (the insulin receptor, growth-factor
receptors), ion-channel receptors (the nicotinic acetylcholine
receptor), and intracellular receptors that bind lipid-soluble ligands
(steroid hormone receptors). The GPCRs are the most diverse class
and the target of approximately $30\,\%$ of currently approved
pharmaceuticals (current as of 2024).

\begin{archetype}[Control and scaling at the molecular machine]
The protein machines of the cell exhibit the control and scaling
archetypes. The control archetype: enzymes are controlled by
allosteric regulators and by covalent modification with response
times of seconds to minutes; ion channels are controlled by voltage,
ligand, and mechanical signals with response times of microseconds
to milliseconds; motor proteins are controlled by the local availability
of ATP and by regulatory binding partners. The scaling archetype:
proteins span four orders of magnitude in mass (from short peptides at
$10^3\,\text{Da}$ to large multi-subunit complexes at
$10^7\,\text{Da}$) with consistent engineering principles applying
across the range. The ribosome, at $2.5$ to $4\,\text{MDa}$, performs
the same template-driven polymer synthesis as a single small enzyme
does in catalysing a single reaction; the architecture scales.
\end{archetype}

\begin{estimation}
\textbf{Question.} Estimate the number of myosin heads engaged
simultaneously with actin filaments in a single fully contracted
human biceps muscle.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A human biceps muscle has a physiological cross-section of
approximately $30\,\text{cm}^2 = 3 \times 10^{-3}\,\text{m}^2$ at
adult maximum. The maximum force per cross-sectional area is
approximately $0.3\,\text{MPa}$, giving a maximum tendon force of
$900\,\text{N}$. The force per engaged myosin head is approximately
$5\,\text{pN} = 5 \times 10^{-12}\,\text{N}$.

The number of simultaneously engaged myosin heads is the muscle force
divided by the per-head force:
\[
N_{\text{heads}} = \frac{900}{5 \times 10^{-12}} = 1.8 \times 10^{14}.
\]

Cross-check from sarcomere geometry: a sarcomere is
$2\,\si{\micro\meter}$ long with approximately $300$ myosin heads
projecting from each thick filament; the biceps cross-section
contains approximately $10^9$ thick filaments (areal density of
$\sim 3 \times 10^{11}/\text{m}^2$ times $3 \times 10^{-3}\,\text{m}^2$).
Total heads in the cross-section is $3 \times 10^{11}$. With a typical
duty cycle of $30\,\%$ to $50\,\%$ of heads engaged at any moment, the
engaged-heads count is $\sim 10^{11}$.

\smallskip
\textit{Postmortem.} The two estimates differ by a factor of
$\sim 10^3$. The first estimate assumes that every myosin head in the
muscle contributes simultaneously, which overcounts because only a
single layer of heads at each thick filament is engaged at one
instant (the heads are spaced along the filament length and engage
sequentially) and only $30\,\%$ to $50\,\%$ of those are bound at any
moment. Applying both corrections to the first estimate brings it
within a factor of five of the geometric estimate. The remaining
discrepancy traces to the per-head force, which varies between
$3\,\text{pN}$ in low-load conditions and $10\,\text{pN}$ at peak
isometric contraction. The two estimates agree to within the
uncertainty of those parameters. The exercise illustrates the
order-of-magnitude character of biological scaling: the number of
engaged motor proteins in a macroscopic action is $\sim 10^{11}$, a
number that is large in absolute terms but small relative to the
total myosin-head inventory.
\end{estimation}

\section{Failure: misfolding diseases}\pathtag{core}

Proteins fail by misfolding. Misfolding is a chemically benign event
when a single protein molecule fails to reach its native fold and is
degraded; it becomes pathological when misfolded proteins accumulate,
aggregate, or evade degradation. The chapter names three mechanism
classes of protein-misfolding disease.

\engterm{Amyloid aggregation diseases}. Many proteins can adopt an
\engterm{amyloid} fold, a $\beta$-sheet-rich aggregated state in which
hundreds or thousands of protein molecules stack into long fibres
held together by cross-$\beta$ structure (extended $\beta$ strands
running perpendicular to the fibre axis, hydrogen-bonded into a
continuous sheet). The amyloid fold is thermodynamically more stable
than the native fold for many proteins; once nucleated, amyloid
formation is self-templating and approximately irreversible. The
diseases associated with amyloid aggregation include Alzheimer's
disease (amyloid-$\beta$ plaques and tau tangles), Parkinson's
disease ($\alpha$-synuclein aggregates), type-2 diabetes (islet
amyloid polypeptide), and the systemic amyloidoses (transthyretin
and other circulating proteins). The mechanism by which amyloid
aggregation produces neurodegeneration is the subject of intense
current research: the aggregates themselves, the intermediate
oligomers, and the resulting cell-stress responses have all been
implicated. The engineering reading is that protein quality control
operates against a thermodynamic preference for aggregation that the
cell normally suppresses but that gradually defeats it with age.

\engterm{Prion diseases}. Prion diseases (Creutzfeldt-Jakob disease,
bovine spongiform encephalopathy, scrapie) are a special case of
protein-misfolding disease in which the misfolded protein
(PrP$^{\text{Sc}}$, scrapie-form) acts as a self-replicating template
that converts the normal cellular protein (PrP$^{\text{C}}$,
cellular-form) to the misfolded state. The prion is therefore
infectious: introduction of misfolded prion into a healthy organism
seeds the conversion of the host's own protein. The engineering
reading is unusual: a misfolded protein is the disease agent and the
replicating entity. The model was proposed by Stanley Prusiner in
1982 and confirmed by extensive subsequent work (Nobel Prize 1997).
Prion diseases are rare but uniformly fatal; the engineering
significance is the demonstration that protein conformation alone,
without nucleic-acid information, can be heritable.

\engterm{Sickle cell disease}. Sickle cell disease is caused by a
single point mutation in the $\beta$-globin gene that substitutes
valine for glutamate at position $6$ of the $\beta$-globin chain.
The mutation changes a charged surface residue to a hydrophobic one;
the resulting haemoglobin S can polymerise into long fibres in the
deoxygenated state, distorting the red blood cell into a sickle
shape. The sickled cells obstruct capillaries, causing painful
vaso-occlusive crises and chronic haemolytic anaemia. The molecular
mechanism (a single hydrophobic side chain inserting into a pocket
on a neighbouring haemoglobin molecule) was worked out by Max Perutz
in the 1960s. Sickle cell disease is one of the clearest examples in
medicine of a single amino-acid change producing a complex clinical
phenotype through molecular aggregation.

\engterm{Quality-control failure}. Underlying all three mechanism
classes is the failure of the cell's protein quality-control
machinery: chaperones to refold or hold misfolded proteins, the
ubiquitin-proteasome system to degrade them, and autophagy to
remove aggregates that escape proteasomal degradation. With age,
quality-control capacity declines and misfolded proteins accumulate.
Many of the chronic neurodegenerative diseases of ageing can be
read as failures of this homeostatic machinery rather than as
diseases of single misfolding events. Therapeutic strategies (small
molecules that stabilise the native fold, antibodies that clear
aggregates, chaperone-upregulation drugs) all aim to compensate for
the failing quality-control system.

\begin{principle}[Protein misfolding is a thermodynamic-and-kinetic competition]
The cell maintains its protein pool in the correct fold by a
competition between native folding (kinetically fast but
thermodynamically marginal) and misfolded or aggregated states
(kinetically slow but thermodynamically more stable for many
sequences). The cell's quality-control machinery shifts the
competition toward the native fold by accelerating native folding
(chaperones), degrading misfolded states (proteasome), and clearing
aggregates (autophagy). Failure modes arise when the competition
shifts: aggregation-prone sequences (sickle cell), self-templating
misfolding (prions), and quality-control decline with age
(neurodegeneration) all reflect the same underlying competition lost
in different ways.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Visualise three proteins from PDB and explain function from structure]
The reader downloads three protein structures from the Protein Data
Bank, visualises them in a free molecular-graphics program (PyMOL,
ChimeraX, or Mol*), and writes a short structure-function analysis
of each.

\textbf{Track}. Computational visualisation.
\textbf{Hazard class}: none.

\textbf{Suggested protein set}.

\begin{itemize}[leftmargin=*]
\item An enzyme: lysozyme (PDB 1LYZ), hexokinase (PDB 1HKB), or
  carbonic anhydrase (PDB 1CA2). The choice exhibits the active site
  and one or more catalytic residues.
\item A motor protein: myosin head (PDB 1B7T or 2MYS) or kinesin
  motor domain (PDB 1BG2). The choice exhibits the nucleotide-binding
  pocket and the cytoskeleton-binding interface.
\item A membrane protein: the bacterial potassium channel KcsA (PDB
  1BL8) or the $\beta_2$-adrenergic receptor (PDB 2RH1). The choice
  exhibits the transmembrane region and the selectivity filter or
  ligand-binding pocket.
\end{itemize}

\textbf{Procedure}.

\begin{enumerate}[leftmargin=*]
\item Download each PDB file and load it into the molecular-graphics
  program of your choice.
\item For each structure, identify the secondary-structure elements,
  the protein's domain organisation, and the principal functional
  region (active site, motor head, channel pore).
\item For each protein, write a one-page structure-function analysis
  that explains how the structure supports the function (which
  residues catalyse the reaction, which residues bind the substrate,
  which residues form the gating mechanism).
\item Compare the three proteins: how do the structural strategies
  differ between an enzyme, a motor, and a channel?
\end{enumerate}

\textbf{Deliverable}. A five- to eight-page report with one rendered
image per protein, the three structure-function analyses, and a
closing paragraph on what the three proteins have in common
(despite their different functions) at the level of fold and
mechanism.

\textbf{Time}. Four to eight hours.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Compute the number of distinct amino-acid sequences of length $100$
that use the $20$ standard amino acids. Compare to the estimated number
of distinct proteins encoded by all life on Earth ($\sim 10^{12}$).
\end{exercise}

\begin{exercise}
A typical $\alpha$-helix has $3.6$ residues per turn and rises
$0.54\,\text{nm}$ per turn. Compute the rise per residue. A
$30$-residue helix is what length?
\end{exercise}

\begin{exercise}
An enzyme has $K_M = 10^{-4}\,\text{M}$ and
$V_{\max} = 100\,\si{\micro\meter}/\text{min}$ with $1\,\si{\micro\meter}$
of enzyme. Compute $k_{\text{cat}}$ in $\text{s}^{-1}$ and the
catalytic efficiency $k_{\text{cat}}/K_M$ in $\text{M}^{-1}\text{s}^{-1}$.
\end{exercise}

\begin{exercise}
A myosin head produces $5\,\text{pN}$ of force over a $5\,\text{nm}$
power stroke per ATP hydrolysed. Compute the work per ATP in
$\text{pN}\cdot\text{nm}$ and in $k_B T$ units at
$310\,\si{\kelvin}$.
\end{exercise}

\begin{exercise}
Kinesin takes $8\,\text{nm}$ steps along microtubules at a rate of
approximately $100\,\text{steps}/\text{s}$. Compute the speed of
cargo transport in $\si{\micro\meter}/\text{s}$ and the time required
to transport cargo from a neuron's cell body to a synapse $1\,\text{m}$
away.
\end{exercise}

\begin{exercise}
A protein has folded-state free energy $\Delta G_f = -40\,\text{kJ}/\text{mol}$
relative to the unfolded state at $310\,\si{\kelvin}$. Compute the
ratio of folded to unfolded molecules in solution and the temperature
at which the protein is half-unfolded if the folding entropy is
$+1500\,\text{J}/(\text{mol}\cdot\si{\kelvin})$.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the Michaelis-Menten equation $v = V_{\max}[S]/(K_M + [S])$
from the steady-state assumption on the enzyme-substrate complex.
Identify the limits in which the rate is first-order in $[S]$ and
zero-order in $[S]$.
\end{exercise}

\begin{exercise}
Derive the Hill equation $\theta = [L]^n/(K^n + [L]^n)$ from the
all-or-nothing assumption that a protein binds either zero or $n$
ligand molecules. Compare to the MWC two-state model.
\end{exercise}

\begin{exercise}
Derive the relation between the persistence length of a polymer and
the polymer's bending modulus $\kappa$: $L_p = \kappa/k_B T$. Use
the result to estimate the bending modulus of the protein
$\alpha$-helix from the helix's persistence length of approximately
$80\,\text{nm}$.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the number of distinct proteins in a typical human cell, given
that the cell expresses approximately $11\,000$ to $13\,000$ genes,
and that alternative splicing roughly doubles the number of distinct
proteins per expressed gene.
\end{exercise}

\begin{exercise}
Estimate the time required for a $100$-residue protein to fold by a
random search through conformational space (Levinthal's paradox).
Compare to the observed folding time of $1$ to $1000\,\text{ms}$ for
typical small proteins.
\end{exercise}

\begin{exercise}
Estimate the cost (in ATP) of synthesising the cell's full protein
complement once. Use the rough total protein mass of an
\emph{E.~coli} cell ($1.5 \times 10^{-13}\,\text{g}$), average
protein mass ($40\,\text{kDa}$), and cost per peptide bond
($\sim 4\,\text{ATP}$ equivalents).
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a numerical solution of the Michaelis-Menten ODE system
(enzyme + substrate + intermediate + product). Compare the numerical
steady-state rate to the analytic Michaelis-Menten prediction and
identify the conditions under which the prediction is inaccurate.
\end{exercise}

\begin{exercise}
Simulate a simple two-state allosteric model with substrate-induced
shift between R and T states. Plot the binding-curve sigmoidicity
as a function of the equilibrium constant $L = [T_0]/[R_0]$ between
states in the absence of substrate.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a thought-experiment fusion protein consisting of a
fluorescent reporter and an allosterically regulated binding domain
that detects a specific small molecule with a $K_d$ in the
nanomolar range. Specify the design constraints and the read-out
strategy.
\end{exercise}

\begin{exercise}
Design a buffer system that maintains a purified enzyme at a stable
fold for a year of refrigerated storage. Identify the principal
denaturing pressures (temperature, pH drift, oxidation, freezing) and
the engineering responses to each.
\end{exercise}

\begin{exercise}
Design a drug-screening strategy for an enzyme target whose
crystal structure is available. Specify the in-silico screening
approach (docking, free-energy perturbation), the in-vitro assay,
and the expected hit rate.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
An enzyme that should catalyse the reaction $A \rightarrow B$ in a
cell extract produces $A \rightarrow B$ at $10\,\%$ of the expected
rate. Identify three possible causes (mutation, post-translational
modification, allosteric inhibitor, denaturing condition) and propose
a diagnostic measurement for each.
\end{exercise}

\begin{exercise}
A purified recombinant protein has the expected sequence but binds
its substrate with $100$ times lower affinity than the native
protein from cell extract. Identify three possible causes (missing
post-translational modification, missing cofactor, partial
unfolding, missing binding partner) and propose a measurement to
distinguish them.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A patient's brain biopsy reveals extracellular amyloid-$\beta$
plaques and intracellular tau tangles. Identify the protein-misfolding
mechanism class (Alzheimer's-type amyloid aggregation) and the
chain of events from misfolding to neuronal death.
\end{exercise}

\begin{exercise}
A research bacterial strain produces an engineered protein at the
expected mRNA level but the protein is recovered in insoluble
inclusion bodies rather than in soluble cytoplasm. Identify three
possible engineering responses (lower expression temperature, coexpress
chaperones, refold from inclusion bodies in vitro) and identify the
mechanistic basis of each.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A press release announces that ``AlphaFold has solved the protein
folding problem''. Identify two ways in which the claim is accurate
(structure prediction is now near experimental accuracy for many
single-domain proteins) and two ways in which it is misleading
(dynamics, conformational ensembles, and ligand binding remain
unsolved at the time of writing).
\end{exercise}

\begin{exercise}
A textbook describes the ribosome as a ``protein factory''.
Identify two ways in which the factory metaphor is accurate
(template-driven, ATP-consuming, modular) and two ways in which it
is misleading (the ribosome is itself a protein-RNA complex, and the
factory metaphor obscures the role of RNA catalysis at the peptidyl
transferase centre).
\end{exercise}

\begin{exercise}
A start-up advertises ``rational drug design'' that designs drug
molecules by computer with $90\,\%$ success rate. Identify the
engineering benchmarks the success rate must satisfy (binding
affinity, selectivity, oral bioavailability, toxicity) and write a
one-paragraph reply naming the bottleneck in modern drug discovery
that the advertised success rate does not address.
\end{exercise}

\begin{exercise}
A reader argues that protein engineering by de novo design is now
``solved'' because the Baker lab and others have demonstrated dozens
of designed proteins that fold to the intended structure. Identify
two engineering distinctions the claim glosses over (fold versus
function, monomer versus dynamic complex) and write a one-paragraph
reply situating the de-novo-design programme within the broader
protein-engineering portfolio.
\end{exercise}
