\chapter{Probability theory}
\label{vol02:ch13}

\epigraph{Probability is the very guide of life.}{Bishop Butler,
\emph{The Analogy of Religion} (1736), introduction; quoted as
the working motto of probability theory in Ross,
\emph{A First Course in Probability}
\cite[ch.~1]{text:ross-probability}.}

\chapmeta{Half-life: foundational. Archetypes: uncertainty
(formalised, deepened from Volume I Chapter 4 and Chapter 8).
Exercise target: 40. Project track: simulation with analysis
(a Markov chain of at least $10\,000$ steps with empirical
state-occupancy compared against the analytical stationary
distribution, plus a 1500-word reflection). Hazard class: none.
Reader path default: standard, with sections explicitly tagged.
Named cases: none in this chapter. The 2008 financial-crisis
correlation pattern in section 13.9 is treated at recognition level
as a generic failure of the independence assumption, without a
named-case registry entry.}

\noindent
Volume I Chapter 8 introduced probability through frequency
counting on engineering samples. We now supply the formal
apparatus. A sample space; a measure; a random variable; an
expectation. The vocabulary lets us state what an inference, an
estimator, and a stochastic process are without depending on the
sample at hand. The reader who finishes this chapter has the
language Volume II Chapter 14 will use to do statistical inference,
Volume IX will use to do queueing theory, and Volume X will use to
do reliability analysis.

\section{Sample spaces and events}\label{sec:vol02:ch13:spaces}\pathtag{core}

A probability model has three pieces. A set of possible outcomes; a
collection of subsets of that set we call events; an assignment of
numbers to those subsets. The numbers obey three rules. The rules
are the foundation on which everything else in this chapter is
built \cite[ch.~1]{text:ross-probability,text:v2ch13-grimmett-stirzaker}.

\subsection{Outcomes, events, axioms}

A \engterm{sample space} $\Omega$ is the set of every distinguishable
outcome of the random experiment under study. A coin toss has
$\Omega = \{H, T\}$. A roll of a six-sided die has $\Omega =
\{1, 2, 3, 4, 5, 6\}$. The lifetime of a single component, modelled
in continuous time, has $\Omega = [0, \infty)$. The choice of sample
space is an engineering choice. A coarser space loses information;
a finer space carries information the model cannot use. Volume I
Chapter 8 made the choice implicitly when it counted defects per
wafer; we now make the choice explicit.

An \engterm{event} $A \subseteq \Omega$ is a subset of the sample
space, interpreted as a statement about the outcome. The event
``the die shows an even number'' is $A = \{2, 4, 6\}$. The event
``the lifetime exceeds $1000\,\text{h}$'' is $A = (1000, \infty)$.
Events combine by the standard set operations: $A \cup B$ is the
event that $A$ or $B$ (or both) occurs; $A \cap B$ is the event that
both occur; $A^{c} = \Omega \setminus A$ is the event that $A$ does
not occur. Two events are \engterm{disjoint} or \engterm{mutually
exclusive} when $A \cap B = \emptyset$.

A collection $\mathcal{F}$ of events that is closed under
complementation and countable union, and contains $\Omega$, is a
\engterm{$\sigma$-algebra}. For finite or countable $\Omega$ we may
take $\mathcal{F}$ to be the power set. For $\Omega = \R$ we take
$\mathcal{F}$ to be the Borel $\sigma$-algebra, generated by the
open intervals; the engineer rarely meets a set on the real line
that is not Borel, and we proceed without dwelling on the
construction \cite[appendix]{text:wasserman2004}.

A \engterm{probability measure} $\Pr$ on $(\Omega, \mathcal{F})$ is
a function $\Pr : \mathcal{F} \to [0, 1]$ satisfying three axioms,
due to Kolmogorov:

\begin{itemize}[leftmargin=*]
  \item (Non-negativity) $\Pr(A) \geq 0$ for every event $A$.
  \item (Normalisation) $\Pr(\Omega) = 1$.
  \item (Countable additivity) For any countable collection of
    pairwise-disjoint events $A_{1}, A_{2}, \ldots$,
    $\Pr\left(\bigcup_{i=1}^{\infty} A_{i}\right) =
    \sum_{i=1}^{\infty} \Pr(A_{i})$.
\end{itemize}

The triple $(\Omega, \mathcal{F}, \Pr)$ is a \engterm{probability
space}. Every probabilistic claim in this chapter sits inside one.

\subsection{Consequences of the axioms}

The axioms are sparing; their consequences are not. Direct
manipulation gives the rules an engineer uses every day:

\begin{itemize}[leftmargin=*]
  \item $\Pr(\emptyset) = 0$.
  \item $\Pr(A^{c}) = 1 - \Pr(A)$.
  \item Monotonicity: $A \subseteq B$ implies $\Pr(A) \leq \Pr(B)$.
  \item Inclusion-exclusion: $\Pr(A \cup B) = \Pr(A) + \Pr(B) -
    \Pr(A \cap B)$.
  \item Continuity from below: if $A_{1} \subseteq A_{2} \subseteq
    \cdots$ and $A = \bigcup A_{i}$, then $\Pr(A_{i}) \to \Pr(A)$.
  \item Boole's inequality (the union bound):
    $\Pr\left(\bigcup_{i=1}^{n} A_{i}\right) \leq \sum_{i=1}^{n}
    \Pr(A_{i})$.
\end{itemize}

Each of the six is a one-line derivation from the axioms; the
exercises ask the reader to do them.

\begin{mastery}
\textbf{Inclusion-exclusion for $n$ events.} The two-event formula
generalises. For events $A_{1}, \ldots, A_{n}$,
\[
\Pr\!\left(\bigcup_{i=1}^{n} A_{i}\right) =
\sum_{i} \Pr(A_{i})
- \sum_{i < j} \Pr(A_{i} \cap A_{j})
+ \sum_{i < j < k} \Pr(A_{i} \cap A_{j} \cap A_{k})
- \cdots
+ (-1)^{n+1} \Pr\!\left(\bigcap_{i=1}^{n} A_{i}\right).
\]
The alternating partial sums bracket the truth: stopping after the
positive single-event terms gives the union bound (an upper bound),
stopping after the pairwise subtraction gives a lower bound, and so
on. These are the \engterm{Bonferroni inequalities}. They let an
engineer bound a joint probability from above and below using only
the low-order joint terms, which are the ones a model can usually
supply. A worked instance: the probability that at least one of
three independent subsystems with failure probabilities $0.02$,
$0.03$, $0.05$ fails is, exactly,
$1 - (0.98)(0.97)(0.95) = 1 - 0.90307 = 0.09693$; the union bound
gives $0.10$, and the two-term Bonferroni lower bound gives
$0.10 - (0.0006 + 0.0010 + 0.0015) = 0.0969$, already correct to
three figures because the pairwise terms are tiny.
\end{mastery}

The union bound deserves a flag. It is loose, often by orders of magnitude when the events
overlap heavily, and it is the engineer's first reach when a
correlated joint event is being upper-bounded by the sum of its
marginals. A safety analysis that uses the union bound on highly
correlated failure modes overestimates the failure probability; a
safety analysis that uses the sum-of-marginals approximation
\emph{instead of} the union bound, treating the events as
independent, will go the other way. Section 13.8 names this latter
error as the dominant misuse pattern in finance and risk.

\subsection{Counting and equally likely outcomes}

For finite $\Omega$ with $|\Omega|$ outcomes that are physically
indistinguishable in their generation (a fair coin, a fair die, a
well-shuffled deck), the probability of an event $A$ reduces to a
counting problem,
\[
\Pr(A) = \frac{|A|}{|\Omega|}.
\]
Counting then proceeds by the standard combinatorial machinery: the
multiplication principle, permutations $n!/(n-k)!$, combinations
$\binom{n}{k} = n!/(k!(n-k)!)$, and the inclusion-exclusion formula
generalised to $n$ events \cite[ch.~1]{text:ross-probability,text:v2ch13-feller-vol1}.
We take these as Volume II Chapter 17 (discrete mathematics) develops
them; the reader who needs a refresher is referred there. The urn of
Figure~\ref{fig:vol02:ch13:urn-counting} is the working example for
the counting exercises and for the contrast between sampling with
and without replacement.

\input{volumes/02-form/ch13-probability/figures/fig-urn-counting.tex}

\paragraph{Worked example: the birthday coincidence.}
A team of $23$ engineers shares a project room. What is the
probability that two of them share a birthday? The estimate that
most people carry into this problem is wrong by a wide margin, and
the discrepancy is instructive. With $365$ possible birthdays and
$23$ people, intuition reaches for a ratio like $23/365 \approx
0.063$ and reports a small chance. The correct calculation counts
the complementary event, all birthdays distinct. The first person
fixes a birthday; the second avoids it with probability $364/365$;
the third avoids the first two with probability $363/365$; and so
on. The probability that all $23$ are distinct is
\[
\Pr(\text{all distinct}) = \prod_{k=0}^{22} \frac{365 - k}{365}
= \frac{365!}{(365 - 23)! \, 365^{23}} \approx 0.4927,
\]
so the probability of at least one shared birthday is
$1 - 0.4927 = 0.5073$, just above one half. The number of
\emph{pairs} among $23$ people is $\binom{23}{2} = 253$, and each
pair coincides with probability $1/365$; the expected number of
coincident pairs is $253/365 \approx 0.69$, which by the Poisson
approximation gives $\Pr(\text{at least one}) \approx 1 - e^{-0.69}
\approx 0.498$, agreeing with the exact count to two figures. The
engineering content is not the party trick. Hash-collision
analysis, the load on a hash table, and the probability that two
randomly generated identifiers collide in a distributed system all
follow the same birthday arithmetic: a collision becomes likely
once the number of items reaches the square root of the number of
slots, not a fixed fraction of it. A $128$-bit identifier space
reaches a $50\,\%$ collision probability at about $2^{64}$ issued
identifiers, which is the design number a systems engineer needs and
the linear intuition would miss by nineteen orders of magnitude.

\paragraph{Worked example: sampling with and without replacement.}
The urn of Figure~\ref{fig:vol02:ch13:urn-counting} holds $N = 50$
components, of which $K = 5$ are defective. An inspector draws
$n = 10$ without replacement. The number of defectives in the draw
follows the \engterm{hypergeometric distribution},
\[
\Pr(X = k) = \frac{\binom{K}{k}\binom{N - K}{n - k}}
{\binom{N}{n}},
\qquad k = 0, 1, \ldots, \min(n, K).
\]
The probability of drawing no defective is
$\binom{5}{0}\binom{45}{10}/\binom{50}{10} =
\binom{45}{10}/\binom{50}{10}$. Evaluating the ratio,
$\binom{45}{10}/\binom{50}{10} = \frac{45!\,40!}{35!\,50!} =
\frac{40 \cdot 39 \cdot 38 \cdot 37 \cdot 36}
{50 \cdot 49 \cdot 48 \cdot 47 \cdot 46} \approx 0.311$. If instead
the inspector samples \emph{with} replacement, each draw is an
independent $\mathrm{Bern}(5/50 = 0.1)$ trial and the count is
$\mathrm{Bin}(10, 0.1)$, giving $\Pr(X = 0) = 0.9^{10} \approx
0.349$. The two answers differ by about four percentage points at
this sample fraction $n/N = 0.2$. The binomial overstates the
chance of a clean sample because, without replacement, each
defective drawn raises the defective fraction among the remainder.
The hypergeometric collapses to the binomial as $N \to \infty$ with
$K/N$ fixed; the finite-population correction factor on the variance
is $(N - n)/(N - 1)$, which an auditor sampling a sizeable fraction
of a finite lot must apply and a textbook example on an infinite
population may ignore.

\begin{mastery}
\textbf{Why the axioms need measure theory.} The countable additivity
axiom looks like a technicality and is the load-bearing one. On a finite
or countable sample space it is harmless: assign a probability to each
outcome and add. On an uncountable space such as $[0, 1]$ it cannot be
satisfied while assigning a probability to \emph{every} subset. Vitali's
construction produces a subset of $[0, 1]$ to which no
translation-invariant countably-additive measure can assign a
consistent value; the same obstruction underlies the Banach-Tarski
paradox. The resolution is to give up measuring every subset and keep
only the \engterm{Borel} (or Lebesgue-measurable) sets, the
$\sigma$-algebra generated by the intervals, which is closed under the
countable operations the axioms require and contains every set an
engineer will ever name. A random variable is then required to be
\engterm{measurable}, meaning the preimage of every Borel set is an
event, precisely so that $\Pr(X \in B)$ is defined for the sets that
carry probability. Expectation is the Lebesgue integral $\E[X] = \int_{\Omega}
X \,\dd\Pr$ against the probability measure, which is why the discrete sum
and the continuous integral of section 13.3 are two cases of one
definition and why a mixed random variable (atom plus density) needs no
special treatment: the single integral handles all three. The machinery
also delivers the convergence theorems (monotone convergence, dominated
convergence, Fatou) that justify interchanging limits and expectations,
the step a derivation quietly takes whenever it differentiates under an
integral sign or passes a limit through an expectation. The engineer
rarely invokes the construction, but it is the reason the symbols
manipulated so freely in the rest of the chapter never contradict each
other.
\end{mastery}

\begin{archetype}[Uncertainty, formalised]
Volume I Chapter 4 carried uncertainty as a single number $u(x)$
beside each measurement. Volume I Chapter 8 carried it as a
distribution shape and a sample-derived parameter. This chapter
carries it as a probability measure on a sample space. The three
treatments are the same archetype at increasing depth. The
measurement uncertainty is the standard deviation of a posited
distribution; the sample standard deviation is an estimator of that
distribution's parameter; the distribution is a probability measure
on the outcome space of the measurement experiment. The reader who
internalises this chain reads a measurement report, a confidence
interval, and a probabilistic safety claim as three views of the
same apparatus.
\end{archetype}

\begin{estimation}
\textbf{Question.} A reliability engineer has five components in
series, each with independent failure probability $0.01$ over a
mission. What is the probability that the system survives?

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A first-order estimate uses the union bound on the failure event:
$\Pr(\text{any failure}) \leq 5 \cdot 0.01 = 0.05$, so the survival
probability is at least $0.95$. A more careful calculation, using
independence, is $\Pr(\text{survive}) = (1 - 0.01)^{5} = 0.99^{5}
\approx 0.951$. The union bound is tight here because the failure
probabilities are small; it is loose when the failure probabilities
are large or the events are heavily correlated. With ten components
at $p = 0.1$ each, the union bound gives an upper bound of $1.0$ on
the failure probability, the independence calculation gives
$1 - 0.9^{10} \approx 0.65$, and the truth depends on the
correlation structure. The estimate that the engineer should take
forward from this chapter is that the union bound is reliable and
loose; independence is strong and often false; the true number sits
between them.
\end{estimation}

\section{Conditional probability and Bayes' rule}\label{sec:vol02:ch13:bayes}\pathtag{core}

Volume I Chapter 8 §8.5 introduced Bayes' rule with a worked
contingency table. We now state the formal apparatus
\cite[ch.~3]{text:ross-probability}.

\subsection{Conditioning}

For events $A, B$ with $\Pr(B) > 0$, the \engterm{conditional
probability of $A$ given $B$} is
\[
\Pr(A \mid B) = \frac{\Pr(A \cap B)}{\Pr(B)}.
\]
The definition treats $\Pr(\cdot \mid B)$ as a probability measure
in its own right on the restricted sample space $B$. The three
Kolmogorov axioms hold for $\Pr(\cdot \mid B)$, with the
normalisation $\Pr(B \mid B) = 1$ replacing $\Pr(\Omega) = 1$.

Two events $A$ and $B$ are \engterm{independent} when
\[
\Pr(A \cap B) = \Pr(A) \Pr(B).
\]
Equivalently, when $\Pr(B) > 0$, $\Pr(A \mid B) = \Pr(A)$:
conditioning on $B$ does not change the probability of $A$.
Independence is a property of the joint measure, not of the events
in isolation; an event that is independent of $B$ may fail to be
independent of $C$. Pairwise independence does not imply mutual
independence: three events $A_{1}, A_{2}, A_{3}$ may satisfy
$\Pr(A_{i} \cap A_{j}) = \Pr(A_{i})\Pr(A_{j})$ for every pair while
$\Pr(A_{1} \cap A_{2} \cap A_{3}) \neq \Pr(A_{1})\Pr(A_{2})\Pr(A_{3})$
\cite[ch.~3]{text:ross-probability}. The exercises ask the reader to
construct a counterexample on a fair-coin sample space.

\paragraph{Worked example: conditioning changes the odds.} Two
machines, $M_{1}$ and $M_{2}$, share a maintenance crew. On any
day, $M_{1}$ is down with probability $0.1$ and $M_{2}$ is down with
probability $0.1$. If the two were independent, both would be down
on $0.1 \cdot 0.1 = 0.01$ of days. Suppose instead a shared power
event, occurring on $0.05$ of days, takes both down at once, and
absent that event each machine is independently down on $0.0526$ of
days (so the marginals still come to $0.1$). Then
$\Pr(\text{both down}) = 0.05 + (1 - 0.05)(0.0526)(0.0526)
\approx 0.05 + 0.0026 = 0.0526$, more than five times the
independence figure. The conditional probability that $M_{2}$ is
down given $M_{1}$ is down is $\Pr(\text{both})/\Pr(M_{1}\text{
down}) = 0.0526/0.1 = 0.526$, against the marginal $0.1$:
observing one machine down raises the probability the other is down
by a factor of five. The shared crew and shared power are the common
cause that breaks independence, and the worked numbers show how
badly the independence figure understates the joint downtime that a
spares-and-staffing plan must cover.

\subsection{The chain rule and the law of total probability}

The definition rearranged is the \engterm{multiplication rule},
$\Pr(A \cap B) = \Pr(A \mid B) \Pr(B)$. Iterated, it becomes the
\engterm{chain rule},
\[
\Pr(A_{1} \cap A_{2} \cap \cdots \cap A_{n}) =
\Pr(A_{1}) \Pr(A_{2} \mid A_{1}) \Pr(A_{3} \mid A_{1} \cap A_{2})
\cdots \Pr(A_{n} \mid A_{1} \cap \cdots \cap A_{n-1}).
\]
Every joint probability factors as a product of conditionals along
some ordering of the events. Different orderings give different
factorisations; the engineer chooses the one in which the
conditionals are most directly available from the model.

A \engterm{partition} of $\Omega$ is a collection $\{B_{1}, \ldots,
B_{k}\}$ of pairwise-disjoint events whose union is $\Omega$. The
\engterm{law of total probability} states that for any event $A$,
\[
\Pr(A) = \sum_{i=1}^{k} \Pr(A \mid B_{i}) \Pr(B_{i}).
\]
The total probability is a weighted average of the conditionals,
with the weights given by the prior probability of each cell of the
partition. In words: every uncertain outcome decomposes by cases.

\paragraph{Worked example: defects from three production lines.}
A plant runs three lines. Line $A$ produces $50\,\%$ of the
output at a defect rate of $1\,\%$; line $B$ produces $30\,\%$ at
$3\,\%$; line $C$ produces $20\,\%$ at $5\,\%$. A unit pulled at
random from the combined output is defective. The marginal defect
rate, by total probability, is
\[
\Pr(D) = 0.50 \cdot 0.01 + 0.30 \cdot 0.03 + 0.20 \cdot 0.05
= 0.005 + 0.009 + 0.010 = 0.024,
\]
so $2.4\,\%$ of units shipped are defective. The follow-up question
is the one Bayes answers: given that a returned unit is defective,
which line most likely made it? The posteriors are
\[
\Pr(A \mid D) = \frac{0.005}{0.024} \approx 0.208, \quad
\Pr(B \mid D) = \frac{0.009}{0.024} = 0.375, \quad
\Pr(C \mid D) = \frac{0.010}{0.024} \approx 0.417.
\]
Line $C$, although it makes the smallest share of output, is the
most likely source of any given defect, because its high defect
rate outweighs its low volume. An engineer who allocates
corrective-action effort by output share would target line $A$
first and waste the budget; the posterior says start with line $C$.
The same arithmetic underlies spam filtering, fault localisation,
and supplier-quality scoring.

\subsection{Bayes' rule}

The \engterm{Bayes rule} follows from the multiplication rule
applied two ways:
\[
\Pr(B_{i} \mid A) = \frac{\Pr(A \mid B_{i}) \Pr(B_{i})}
{\sum_{j=1}^{k} \Pr(A \mid B_{j}) \Pr(B_{j})}.
\]
The numerator is the \engterm{likelihood} of $A$ under hypothesis
$B_{i}$ multiplied by the \engterm{prior} on $B_{i}$. The
denominator is the marginal probability of $A$, computed by total
probability. The output is the \engterm{posterior} probability of
$B_{i}$ after observing $A$. Volume I Chapter 8 §8.5 gave the
worked diagnostic-test calculation; the reader who has not absorbed
that example should revisit it now.

\paragraph{Worked example: the diagnostic test and its base rate.}
A non-destructive inspection flags microcracks in turbine blades.
The method has \engterm{sensitivity} $\Pr(+ \mid \text{cracked}) =
0.98$ and \engterm{specificity} $\Pr(- \mid \text{sound}) = 0.95$,
so its false-positive rate is $0.05$. In the fleet under study, the
true prevalence of a cracked blade is $\Pr(\text{cracked}) = 0.004$,
four in a thousand. A blade flags positive. The probability it is
actually cracked is
\[
\Pr(\text{cracked} \mid +)
= \frac{0.98 \cdot 0.004}
{0.98 \cdot 0.004 + 0.05 \cdot 0.996}
= \frac{0.00392}{0.00392 + 0.0498}
= \frac{0.00392}{0.05372} \approx 0.073.
\]
A test that is right $98\,\%$ of the time on cracked blades and
$95\,\%$ of the time on sound ones yields a positive result that is
correct only about $7\,\%$ of the time. The reason sits in the
denominator: the sound population is $249$ times larger than the
cracked population, so even a small false-positive rate generates
far more false alarms than the rare true cracks generate true
alarms. Of every $10\,000$ blades inspected, $40$ are cracked and
about $39$ of those flag; $9960$ are sound and about $498$ of those
flag falsely; the positive count is $537$, of which $39$, or
$7.3\,\%$, are real. Figure~\ref{fig:vol02:ch13:bayes-tree} renders
the same calculation as a tree, with the four leaf counts that
populate the contingency table. The engineering response is not to
abandon the test but to read the posterior correctly: a single
positive on a rare defect warrants a second, mechanism-diverse test
(the sequential design of the exercises), never an immediate
scrap-or-replace decision. Raising the prevalence by pre-screening
on age or duty cycle, so the test is applied to a population where
$\Pr(\text{cracked})$ is $0.05$ rather than $0.004$, lifts the
post-test probability from $7\,\%$ to about $51\,\%$ with no change
to the instrument.

\input{volumes/02-form/ch13-probability/figures/fig-bayes-tree.tex}

\subsection{The base-rate fallacy}

The most stable mistake in applied conditional probability is to
read $\Pr(A \mid B)$ as $\Pr(B \mid A)$. The two are linked by
Bayes' rule and equal only when $\Pr(A) = \Pr(B)$. A medical or
engineering screening test with high sensitivity and high
specificity can still produce a posterior near zero when the prior
is small; the worked example in Volume I Chapter 8 showed a
posterior of about $0.32$ from a $95\,\%$-sensitive, $98\,\%$-specific
test on a $1\,\%$-prior population. The discipline is to write the
contingency table whenever the answer is decision-relevant. The
engineer who reasons by the test characteristics alone, without the
prior, will systematically overestimate the rate of true positives
on rare events. Section 13.8 returns to this pattern as a class of
correlated-failure misjudgements.
Figure~\ref{fig:vol02:ch13:bayes-schematic} lays out the
contingency table for the running example so the reader can see
which cell carries the dominant term in the denominator. The
companion script
\repopath{volumes/02-form/ch13-probability/code/bayes_table.py}
prints the same table for any prior, sensitivity, and specificity
combination of the reader's choice.

\input{volumes/02-form/ch13-probability/figures/fig-bayes-schematic.tex}

\begin{mastery}
\textbf{Bayes in odds form.} Dividing the posterior for a
hypothesis $H$ by the posterior for its complement removes the
denominator and exposes the structure:
\[
\underbrace{\frac{\Pr(H \mid E)}{\Pr(H^{c} \mid E)}}
_{\text{posterior odds}}
=
\underbrace{\frac{\Pr(E \mid H)}{\Pr(E \mid H^{c})}}
_{\text{likelihood ratio}}
\times
\underbrace{\frac{\Pr(H)}{\Pr(H^{c})}}
_{\text{prior odds}}.
\]
Each independent piece of evidence multiplies the odds by its own
likelihood ratio, so on the log scale evidence adds. For the
turbine-blade test, the likelihood ratio of a positive is
$0.98/0.05 = 19.6$ and the prior odds are
$0.004/0.996 \approx 0.00402$, giving posterior odds
$19.6 \times 0.00402 \approx 0.0788$, a posterior probability of
$0.0788/(1 + 0.0788) \approx 0.073$, matching the direct
calculation. A second independent positive multiplies the odds by
$19.6$ again, reaching $1.54$, a posterior of about $0.61$. The
odds form is the natural home for sequential updating and the
cleanest way to see that one strong test rarely overturns a very
small prior, while two independent moderate tests can.
\end{mastery}

\paragraph{The prosecutor's fallacy.} The base-rate error has a
twin that misreads the same two conditionals in the legal and
forensic setting. A forensic match has a random-match probability
$\Pr(\text{match} \mid \text{innocent}) = 10^{-6}$. The prosecutor
states that the probability the defendant is innocent is therefore
$10^{-6}$, conflating $\Pr(\text{match} \mid \text{innocent})$ with
$\Pr(\text{innocent} \mid \text{match})$. In a city of $10^{6}$
plausible suspects, the expected number of innocent matches is
about $1$, so a lone match leaves the posterior probability of guilt
near $1/2$, not $1 - 10^{-6}$. The engineering analogue arises
whenever a rare diagnostic signature is read as near-certain
identification of a fault across a large population of components.
The discipline is identical to the base-rate discipline: write the
two conditionals separately and let Bayes connect them.

Events $A$ and $B$ are \engterm{conditionally independent given
$C$} when $\Pr(A \cap B \mid C) = \Pr(A \mid C) \Pr(B \mid C)$.
Conditional independence does not imply unconditional independence,
and unconditional independence does not imply conditional
independence. Two sensors with independent measurement noise
become correlated when they are conditioned on a shared upstream
fault; two diseases with correlated prevalence may be conditionally
independent given a hidden common cause. The Bayesian-network
literature \cite[ch.~11]{text:wasserman2004} formalises which
conditionings preserve and which destroy independence. We use the
notion in section 13.5 (independence and correlation) and again in
Volume IX (control) where state-space models exploit conditional
independence to make inference tractable.

\subsection{Recursive Bayes and sensor fusion}

Bayes' rule runs once on a static contingency table; it runs forever on
a stream. When evidence $E_{1}, E_{2}, \ldots$ arrives in sequence and
each piece is conditionally independent of the others given the
hypothesis, the posterior after $E_{1}$ becomes the prior for $E_{2}$,
and the update is recursive,
\[
\Pr(H \mid E_{1:t}) \propto \Pr(E_{t} \mid H)\,
\Pr(H \mid E_{1:t-1}),
\]
which is the odds-form sequential update of the Bayes mastery box read
as a running computation rather than a one-step formula. The recursion
is the heart of \engterm{sensor fusion}: each sensor supplies a
likelihood $\Pr(E_{t} \mid H)$, and the fused belief is the prior
multiplied by every sensor's likelihood, with the normalisation deferred
to the end. Two sensors that disagree pull the posterior toward whichever
carries the sharper likelihood (the smaller measurement variance), and a
sensor with a flat likelihood contributes nothing, which is the
quantitative statement of how a fusion system weights its inputs by
their precision.

\paragraph{Worked example: fusing two range estimates.} A robot
estimates its distance to a wall with two sensors. A sonar reports
$d_{1} = 2.10\,\text{m}$ with measurement standard deviation $\sigma_{1}
= 0.20\,\text{m}$; a laser reports $d_{2} = 2.02\,\text{m}$ with
$\sigma_{2} = 0.05\,\text{m}$. Treating each reading as a Gaussian
likelihood for the true distance and starting from a flat prior, the
fused posterior is Gaussian with precision (inverse variance) the sum of
the two precisions,
\[
\frac{1}{\sigma^{2}} = \frac{1}{\sigma_{1}^{2}} +
\frac{1}{\sigma_{2}^{2}} = \frac{1}{0.04} + \frac{1}{0.0025}
= 25 + 400 = 425\,\text{m}^{-2},
\]
so $\sigma = 0.0485\,\text{m}$, and the fused mean is the
precision-weighted average,
\[
\hat d = \sigma^{2}\!\left(\frac{d_{1}}{\sigma_{1}^{2}} +
\frac{d_{2}}{\sigma_{2}^{2}}\right)
= \frac{25 \cdot 2.10 + 400 \cdot 2.02}{425}
= \frac{52.5 + 808}{425} = 2.025\,\text{m}.
\]
The fused estimate sits almost on top of the laser reading because the
laser is sixteen times more precise (precision goes as $1/\sigma^{2}$),
and the fused uncertainty $0.0485\,\text{m}$ is tighter than either
sensor alone: combining independent measurements always sharpens the
estimate, the precision-addition rule that the Kalman filter applies at
every step and that the bivariate-Gaussian conditioning of section 13.5
will derive as a special case. The fusion is honest only while the two
errors are independent; a common bias (both sensors mis-mounted, both
reading off a fogged surface) couples the errors, the precision-addition
overstates the gain, and the result is the independence-assumed-where-
there-was-none failure that section 13.9 names.

\begin{mastery}
\textbf{From Kalman to particle filters.} The recursive Bayes update is
exact and closed-form only when the prior, the likelihood, and the
state dynamics are all linear-Gaussian; then the posterior stays
Gaussian and the recursion is the \engterm{Kalman filter}, carrying a
mean and a covariance forward through a predict step (propagate the
state, inflate the covariance by the process noise) and an update step
(the precision-weighted fusion of the worked example, generalised to
vectors via the Schur complement of section 13.5). When the dynamics are
non-linear or the noise is non-Gaussian, the posterior has no
finite-parameter form, and the \engterm{particle filter} represents it
by a swarm of weighted samples. Each particle is a candidate state; the
predict step pushes every particle through the (possibly non-linear)
dynamics with a random kick; the update step multiplies each particle's
weight by the likelihood of the new observation; and a
\engterm{resampling} step periodically discards low-weight particles and
duplicates high-weight ones to keep the swarm where the posterior mass
is. The swarm is a Monte Carlo representation of the posterior, so its
accuracy obeys the $1/\sqrt{n}$ wall in the number of particles, and its
failure mode is \engterm{weight degeneracy}: without resampling, one
particle accumulates nearly all the weight and the effective sample size
collapses to one. The particle filter is the tool when a robot localises
on a non-linear map, when a tracker follows a manoeuvring target, and
when a process model is too non-linear for the Kalman linearisation;
Volume IX develops both, and the conceptual seed is the recursive Bayes
update of this section run on a sampled, rather than parametric,
representation of belief.
\end{mastery}

\section{Random variables, expectation, variance}\pathtag{core}

A random variable $X$ on a probability space $(\Omega, \mathcal{F},
\Pr)$ is a measurable function $X : \Omega \to \R$. The reader who
has not met the measurability condition may read it as: for every
real $a$, the set $\{\omega \in \Omega : X(\omega) \leq a\}$ is an
event, so the probability $\Pr(X \leq a)$ is defined. The technical
content is that we cannot pull arbitrary subsets of $\R$ back to
$\Omega$ and expect them to have probabilities; we can pull back
the Borel sets, which is enough for engineering work
\cite[ch.~4]{text:wasserman2004}.

\subsection{Distribution functions}

The \engterm{cumulative distribution function} (CDF) of $X$ is
\[
F_{X}(x) = \Pr(X \leq x), \qquad x \in \R.
\]
The CDF is non-decreasing, right-continuous, satisfies $F_{X}(-\infty)
= 0$, and $F_{X}(+\infty) = 1$. Every random variable has a CDF; the
CDF determines the distribution. Probabilities of intervals come
straight from the CDF: $\Pr(a < X \leq b) = F_{X}(b) - F_{X}(a)$.

A random variable is \engterm{discrete} when it takes values in a
countable set $\{x_{1}, x_{2}, \ldots\}$. The \engterm{probability
mass function} (PMF) is $p_{X}(x_{i}) = \Pr(X = x_{i})$, satisfying
$p_{X}(x_{i}) \geq 0$ and $\sum_{i} p_{X}(x_{i}) = 1$.

A random variable is (absolutely) \engterm{continuous} when its CDF
admits a density: a non-negative function $f_{X}$ such that
\[
F_{X}(x) = \int_{-\infty}^{x} f_{X}(t) \,\dd t.
\]
The density is the \engterm{probability density function} (PDF). It
satisfies $f_{X}(x) \geq 0$ and $\int_{-\infty}^{\infty} f_{X}(x)
\,\dd x = 1$. The PDF at a point is not a probability; only its
integral over a set is. The probability that $X$ lies in $[a, b]$ is
$\int_{a}^{b} f_{X}(x) \,\dd x$.

A random variable need be neither discrete nor continuous. Mixtures
arise routinely: the time a system spends in a queue is zero with
positive probability and continuous on $(0, \infty)$.

\subsection{Expectation}

The \engterm{expectation} (or \engterm{mean}) of a discrete random
variable is
\[
\E[X] = \sum_{i} x_{i} \, p_{X}(x_{i}),
\]
when the sum converges absolutely. For a continuous random variable,
\[
\E[X] = \int_{-\infty}^{\infty} x \, f_{X}(x) \,\dd x,
\]
when the integral converges absolutely. The expectation is the
centre of mass of the distribution. Some random variables have no
expectation (the Cauchy distribution, with density
$f(x) = 1/(\pi(1 + x^{2}))$, integrates to one but $\int |x| f(x)
\,\dd x = \infty$); the engineer who fits a Cauchy and then
estimates a mean by the sample average is computing a quantity with
no population analogue.

For any function $g : \R \to \R$ for which the relevant sum or
integral converges absolutely,
\[
\E[g(X)] = \sum_{i} g(x_{i}) p_{X}(x_{i})
\quad\text{or}\quad
\E[g(X)] = \int_{-\infty}^{\infty} g(x) f_{X}(x) \,\dd x.
\]
This is the \engterm{law of the unconscious statistician}. The
reader does not need the distribution of $g(X)$ to compute its
expectation; the distribution of $X$ suffices.

Expectation is linear: for random variables $X, Y$ on the same
space and constants $a, b$,
\[
\E[aX + bY] = a\E[X] + b\E[Y].
\]
Linearity holds without any independence assumption. It is the most
useful single fact in probability theory and the cleanest tool the
engineer has for computing means of complicated sums.

\paragraph{Worked example: expected coupons.}
A process delivers, on each cycle, one of $n$ distinct calibration
constants uniformly at random. How many cycles, on average, until
all $n$ have been seen at least once? Direct enumeration of the
waiting time is hard; linearity makes it easy. Let $T_{i}$ be the
number of additional cycles to collect a new constant once $i - 1$
distinct ones are already in hand. The chance a given cycle is new
is $(n - i + 1)/n$, so $T_{i} \sim \mathrm{Geom}((n - i + 1)/n)$
with mean $n/(n - i + 1)$. The total wait is $T = \sum_{i=1}^{n}
T_{i}$, and by linearity
\[
\E[T] = \sum_{i=1}^{n} \frac{n}{n - i + 1}
= n \sum_{j=1}^{n} \frac{1}{j} = n H_{n}
\approx n(\ln n + 0.577),
\]
where $H_{n}$ is the $n$th harmonic number. For $n = 50$ constants,
$\E[T] \approx 50(\ln 50 + 0.577) \approx 50 \cdot 4.49 \approx 225$
cycles. The last few constants dominate the wait: collecting the
final one alone takes $n$ cycles on average. The coupon-collector
result sets expectations for coverage testing, cache-warming, and
any sampling scheme that must touch every category of a finite set.

\subsection{Variance and standard deviation}

The \engterm{variance} of a random variable with finite mean
$\mu = \E[X]$ is
\[
\Var(X) = \E[(X - \mu)^{2}] = \E[X^{2}] - \mu^{2},
\]
when finite. The variance measures spread around the mean. Its
square root is the \engterm{standard deviation} $\sigma_{X} =
\sqrt{\Var(X)}$, which has the same units as $X$.

Variance is not linear. For constants $a, b$,
\[
\Var(aX + b) = a^{2} \Var(X).
\]
The constant $b$ shifts the distribution without changing its
spread. The constant $a$ scales the spread by $|a|$, which appears
squared in the variance because variance is in squared units.

\paragraph{Worked example: variance of a discrete random variable.}
A shipping container holds parts whose count of surface blemishes
$X$ takes values $0, 1, 2, 3$ with probabilities $0.5, 0.3, 0.15,
0.05$. The mean is
\[
\E[X] = 0(0.5) + 1(0.3) + 2(0.15) + 3(0.05) = 0.75.
\]
The second moment is
\[
\E[X^{2}] = 0(0.5) + 1(0.3) + 4(0.15) + 9(0.05) = 1.35,
\]
so the variance is $\Var(X) = 1.35 - 0.75^{2} = 1.35 - 0.5625 =
0.7875$ and the standard deviation is $\sigma_{X} \approx 0.887$
blemishes. A specification that allows at most one blemish accepts
the event $\{X \leq 1\}$ with probability $0.8$; the engineer who
reports only the mean of $0.75$ hides that one container in five
fails the limit. The mean and the variance together, or better the
full mass function, carry the acceptance information that the mean
alone does not.

For two random variables, the variance of a sum is
\[
\Var(X + Y) = \Var(X) + \Var(Y) + 2\Cov(X, Y),
\]
where $\Cov(X, Y) = \E[(X - \E X)(Y - \E Y)]$ is the
\engterm{covariance}. Covariance is the off-diagonal term that
linearity of expectation does not see. When $X$ and $Y$ are
uncorrelated, $\Cov(X, Y) = 0$ and the variance of the sum reduces
to the sum of variances. When $X$ and $Y$ are positively
correlated, the variance of the sum exceeds the sum of variances.
Section 13.5 develops the consequences for portfolios, sensors, and
parallel components.

\subsection{Higher moments and quantiles}

The $k$th \engterm{moment} of $X$ is $\E[X^{k}]$, when defined. The
$k$th \engterm{central moment} is $\E[(X - \mu)^{k}]$. The third
central moment, normalised by $\sigma^{3}$, is the \engterm{skewness},
a unitless measure of asymmetry. The fourth, normalised by
$\sigma^{4}$, is the \engterm{kurtosis}, a unitless measure of tail
weight.

For continuous $X$ with continuous CDF, the \engterm{$q$-quantile}
is the value $x_{q}$ such that $F_{X}(x_{q}) = q$. The median is
$x_{0.5}$. The 95th percentile is $x_{0.95}$. Quantiles are
preserved under monotone transformations: if $g$ is increasing,
the $q$-quantile of $g(X)$ is $g(x_{q})$. The mean is not preserved
under monotone transformations, which is one reason quantiles are
the right tool for skewed engineering data.

\begin{mastery}
\textbf{Moment generating functions.} The \engterm{moment
generating function} (MGF) of $X$ is $M_{X}(t) = \E[e^{tX}]$ where
the expectation exists for $t$ in a neighbourhood of zero.
Differentiating and setting $t = 0$ recovers the moments:
$M_{X}'(0) = \E[X]$, $M_{X}''(0) = \E[X^{2}]$, and in general
$M_{X}^{(k)}(0) = \E[X^{k}]$. The MGF of a sum of independent
variables is the product of their MGFs, which turns convolution
into multiplication and is the cleanest route to the distribution
of a sum. For $X \sim \mathcal{N}(\mu, \sigma^{2})$,
$M_{X}(t) = \exp(\mu t + \tfrac{1}{2}\sigma^{2} t^{2})$; the product
of two such MGFs is again of this form with $\mu_{1} + \mu_{2}$ and
$\sigma_{1}^{2} + \sigma_{2}^{2}$, which proves that independent
Gaussians sum to a Gaussian. For $X \sim \mathrm{Pois}(\lambda)$,
$M_{X}(t) = \exp(\lambda(e^{t} - 1))$, and the product form shows
that independent Poissons sum to a Poisson with added rates. When
the MGF fails to exist (heavy-tailed laws), the
\engterm{characteristic function} $\E[e^{itX}]$, which always
exists, replaces it; it is the tool behind the proof of the central
limit theorem.
\end{mastery}

\subsection{Markov and Chebyshev inequalities}

The \engterm{Markov inequality} says that for a non-negative random
variable $X$ with finite mean and any $a > 0$,
\[
\Pr(X \geq a) \leq \frac{\E[X]}{a}.
\]
The proof is one line: $a \cdot \mathbf{1}\{X \geq a\} \leq X$,
take expectations, divide by $a$. The bound is loose for thin-tailed
distributions and tight for distributions concentrated on $\{0, a\}$.

The \engterm{Chebyshev inequality}, applied to $(X - \mu)^{2}$ with
$a = k^{2}\sigma^{2}$, gives
\[
\Pr(|X - \mu| \geq k\sigma) \leq \frac{1}{k^{2}}.
\]
For $k = 2$: at most a quarter of the mass lies more than two
standard deviations from the mean. For $k = 3$: at most one ninth.
These bounds hold for every distribution with finite variance,
which is their value: an engineer who knows only $\mu$ and $\sigma$
already has a calibrated tail bound. The bounds are loose; for a
Gaussian the true two-$\sigma$ probability is about $0.046$, against
the Chebyshev bound of $0.25$. A working engineer reaches for
Chebyshev when distributional assumptions are weak and for the
relevant tabulated tail when the distribution is named.

\subsection{Functions of random variables and error propagation}

A measured quantity feeds a calculation, and the uncertainty in the
input becomes uncertainty in the output. The probabilistic question
is the distribution of $Y = g(X)$ given the distribution of $X$.
For a monotone, differentiable $g$ with inverse $h = g^{-1}$, the
density transforms by the \engterm{change-of-variables} formula,
\[
f_{Y}(y) = f_{X}(h(y)) \left| h'(y) \right|.
\]
The Jacobian factor $|h'(y)|$ accounts for the way $g$ stretches or
compresses the probability mass. As a worked instance, if
$X \sim \mathrm{Unif}(0, 1)$ and $Y = -\ln X / \lambda$, then
$h(y) = e^{-\lambda y}$, $|h'(y)| = \lambda e^{-\lambda y}$, and
$f_{Y}(y) = \lambda e^{-\lambda y}$ for $y \geq 0$: a uniform fed
through a negative logarithm produces an exponential. This is the
inverse-transform sampling identity that every random-number
generator relies on, and the project's chain simulation uses it
implicitly.

When the full distribution of $Y$ is not needed, the first-order
\engterm{delta method} propagates mean and variance through a
smooth $g$. Expanding $g$ to first order around $\mu_{X} = \E[X]$,
\[
\E[g(X)] \approx g(\mu_{X}), \qquad
\Var(g(X)) \approx \left[ g'(\mu_{X}) \right]^{2} \Var(X).
\]
For a function of several independent inputs $Y = g(X_{1}, \ldots,
X_{m})$, the variances add through the partial derivatives,
\[
\Var(Y) \approx \sum_{i=1}^{m}
\left( \frac{\partial g}{\partial x_{i}} \right)^{2}
\Var(X_{i}),
\]
evaluated at the means. This is the \engterm{Gaussian error
propagation} formula, the probabilistic statement of the
combined-standard-uncertainty rule that Volume I Chapter 4
introduced for measurement uncertainty. When the inputs are
correlated, the cross terms $2 \sum_{i < j} (\partial g /
\partial x_{i})(\partial g / \partial x_{j}) \Cov(X_{i}, X_{j})$
return, and ignoring them repeats the independence error that
section 13.9 names.
Figure~\ref{fig:vol02:ch13:error-propagation} shows the geometry:
the local slope of $g$ sets the amplification of the input spread.

\input{volumes/02-form/ch13-probability/figures/fig-error-propagation.tex}

\paragraph{Worked example: propagating uncertainty into power
dissipation.} A resistor carries current $I = 2.0 \pm 0.05\,
\text{A}$ (mean and standard deviation) and has resistance
$R = 50 \pm 1\,\Omega$, the two measured independently. The
dissipated power is $P = I^{2} R$. The mean power is approximately
$\mu_{I}^{2} \mu_{R} = 4.0 \cdot 50 = 200\,\text{W}$. The partial
derivatives at the means are $\partial P / \partial I = 2 I R =
2 \cdot 2 \cdot 50 = 200$ and $\partial P / \partial R = I^{2} =
4.0$. The propagated variance is
\[
\Var(P) \approx (200)^{2}(0.05)^{2} + (4.0)^{2}(1)^{2}
= 100 + 16 = 116\,\text{W}^{2},
\]
so $\sigma_{P} \approx 10.8\,\text{W}$, a relative uncertainty of
about $5.4\,\%$. The current term dominates because power depends on
the square of current; halving the current uncertainty would cut
the total uncertainty far more than halving the resistance
uncertainty. The delta method tells the engineer where to spend the
metrology budget without simulating anything. Its limitation is the
linearisation: for strongly non-linear $g$ or large input
uncertainty, the first-order formula understates the spread and a
Monte Carlo propagation (sampling $I$ and $R$, computing $P$ for
each draw) is the honest fallback. The companion script
\repopath{volumes/02-form/ch13-probability/code/error_propagation.py}
runs both the delta method and the Monte Carlo propagation for this
example so the reader can see where the linearisation holds and where
it parts company with the simulation.

\subsection{Non-monotone transformations and the two-branch Jacobian}

The change-of-variables formula above assumed $g$ monotone. When $g$
folds the line, several input values map to one output value, and the
density of $Y = g(X)$ collects a Jacobian term from each preimage
branch. For a piecewise-monotone $g$ with branches $h_{1}, \ldots,
h_{m}$ inverting $g$ on the disjoint pieces that map onto $y$,
\[
f_{Y}(y) = \sum_{j=1}^{m} f_{X}(h_{j}(y)) \left| h_{j}'(y) \right|.
\]
The square map is the working case. Let $X \sim \mathcal{N}(0, 1)$
and $Y = X^{2}$. For $y > 0$ the two preimages are $h_{\pm}(y) =
\pm\sqrt{y}$, each with $|h_{\pm}'(y)| = 1/(2\sqrt{y})$. The standard
normal density is even, so the two branches contribute equally,
\[
f_{Y}(y) = 2 \cdot \frac{1}{\sqrt{2\pi}} e^{-y/2}
\cdot \frac{1}{2\sqrt{y}}
= \frac{1}{\sqrt{2\pi y}} e^{-y/2}, \qquad y > 0.
\]
This is the chi-square density with one degree of freedom, which is
also $\mathrm{Gamma}(\tfrac12, \tfrac12)$ in the shape-rate parameters
of the distribution catalogue. The derivation earns its place beyond
the algebra: it shows where the sum of squared standardised residuals
in a least-squares fit gets its distribution, and it is the reason a
sum of $k$ squared independent standard normals is $\chi^{2}_{k}$, the
reference law for the variance estimates of Volume II Chapter 14. An
engineer who applies the single-branch formula to a folding transform
loses a factor of two and reports a density that does not integrate to
one, a mistake the normalisation check catches immediately.

\paragraph{Worked example: the ratio of two normals and a heavy
tail.} Transformations can manufacture pathology. Let $X, Y$ be
independent $\mathcal{N}(0, 1)$ and $W = X/Y$. The standard
change-of-variables in two dimensions (introduce $V = Y$, invert,
compute the $2 \times 2$ Jacobian, integrate out $V$) gives the
standard Cauchy density $f_{W}(w) = 1/(\pi(1 + w^{2}))$. Two
well-behaved finite-variance inputs produce an output with no mean
and no variance. The engineering warning is direct: a ratio estimator
formed from a noisy denominator that can approach zero, a gain
computed as output over input near a null, a calibration slope read
off a near-flat response, all inherit the Cauchy tail and defeat the
sample average. The fix is to bound the denominator away from zero by
design or to report a location estimate that resists heavy tails (a
trimmed mean or a median) for which the Cauchy still has a
well-defined centre.

\subsection{Order statistics}

Sorting a sample turns the largest, the smallest, and the median into
random variables in their own right, and reliability, extreme-load,
and acceptance problems live on these. For an iid sample $X_{1},
\ldots, X_{n}$ with common CDF $F$ and density $f$, the
\engterm{order statistics} are the sorted values $X_{(1)} \leq X_{(2)}
\leq \cdots \leq X_{(n)}$. The extremes have the cleanest CDFs,
because $X_{(n)} \leq x$ exactly when every sample point is at most
$x$, and $X_{(1)} > x$ exactly when every point exceeds $x$:
\[
F_{X_{(n)}}(x) = F(x)^{n}, \qquad
F_{X_{(1)}}(x) = 1 - (1 - F(x))^{n}.
\]
Differentiating, the maximum has density $n F(x)^{n-1} f(x)$ and the
minimum has density $n(1 - F(x))^{n-1} f(x)$. The general $k$th order
statistic has density
\[
f_{X_{(k)}}(x) = \frac{n!}{(k-1)!\,(n-k)!}
F(x)^{k-1} (1 - F(x))^{n-k} f(x),
\]
read as a multinomial bookkeeping: one point lands at $x$ (the factor
$f(x)$), $k-1$ fall below, $n-k$ fall above, and the factorial counts
the labelings.

\paragraph{Worked example: the weakest link in a chain.} A chain of
$n = 20$ links each has tensile strength $X_{i} \sim
\mathrm{Exp}(\lambda)$ with $\lambda = 1/50\,\text{kN}^{-1}$, so the
mean link strength is $50\,\text{kN}$. The chain breaks at the
strength of its weakest link, $X_{(1)}$. The minimum of $n$ iid
exponentials is itself exponential with the rates added:
$X_{(1)} \sim \mathrm{Exp}(n\lambda)$, with mean
$1/(n\lambda) = 50/20 = 2.5\,\text{kN}$. The chain is twenty times
weaker on average than a single link, and the weakest-link logic is
why a series system of nominally identical components fails far sooner
than any one component would. The median link strength is $50 \ln 2
\approx 34.7\,\text{kN}$, yet the chain's median strength is
$2.5 \ln 2 \approx 1.7\,\text{kN}$. The same minimum-of-many structure
drives the size effect in brittle materials, where larger specimens
contain more candidate flaws and fail at lower nominal stress, the
phenomenon the Weibull distribution was built to describe.

\paragraph{Worked example: the expected range of a uniform sample.}
Five sensors report a reading uniformly distributed on $[0, 1]$
(after rescaling). The spread between the highest and lowest reading,
the \engterm{sample range} $R = X_{(5)} - X_{(1)}$, is a one-number
summary of disagreement. For $n$ iid $\mathrm{Unif}(0, 1)$ variables
the minimum has mean $1/(n+1)$ and the maximum has mean $n/(n+1)$, so
$\E[R] = (n-1)/(n+1)$. For $n = 5$, $\E[R] = 4/6 \approx 0.667$: five
uniform readings span on average two thirds of the available interval.
The result calibrates the intuition that a handful of samples rarely
covers the full range, and it is the analytic backbone of
range-based control charts, where the average range of small subgroups
estimates the process standard deviation through a tabulated constant
$d_{2}$ that is exactly $\E[R]$ for the standard normal.

\input{volumes/02-form/ch13-probability/figures/fig-order-statistics.tex}

\paragraph{Worked example: the 100-year load.} Civil and environmental
codes specify a \engterm{return period}: the load with a $1$-in-$T$
annual exceedance probability, the so-called $T$-year load. If the
annual maximum load $X$ has CDF $F$, the $T$-year load $x_{T}$ solves
$\Pr(X > x_{T}) = 1 - F(x_{T}) = 1/T$, so $x_{T} = F^{-1}(1 - 1/T)$. The
return period is the reciprocal of the annual exceedance probability,
and the expected waiting time to the first exceedance is, by the
geometric distribution of section 13.4, exactly $T$ years. The name
misleads on purpose: a $100$-year load is not a load that arrives every
hundred years on schedule, but one with probability $0.01$ of being
exceeded in any given year. Over a $50$-year design life the probability
that the $100$-year load is exceeded at least once is $1 - (1 -
0.01)^{50} = 1 - 0.605 = 0.395$, almost two in five, which is the number
a designer needs and the every-hundred-years reading would badly miss.
Demanding only a $10\,\%$ lifetime exceedance probability over $50$
years requires the load with annual exceedance $p$ solving $(1 - p)^{50}
= 0.90$, that is $p \approx 0.0021$, a return period near $475$ years,
which is exactly why seismic codes quote a $475$-year event for a
$10\,\%$-in-$50$-year hazard. The return-period arithmetic is the bridge
from an annual probability to a lifetime risk, and it is the order-
statistic maximum read through the geometric waiting time.

\begin{mastery}
\textbf{Extreme-value theory and the three limit laws.} The central
limit theorem governs sums; a parallel theory governs maxima. The
\engterm{Fisher-Tippett-Gnedenko theorem} states that if the normalised
maximum $(X_{(n)} - b_{n})/a_{n}$ of an iid sample converges to a
non-degenerate law, that law is the \engterm{generalised extreme value}
(GEV) distribution,
\[
G(x) = \exp\!\left( -\left[1 + \xi\,\frac{x - \nu}{\beta}\right]
^{-1/\xi} \right),
\]
with location $\nu$, scale $\beta > 0$, and shape $\xi$. The shape
$\xi$ selects one of three families: $\xi = 0$ (read as the limit) is
the \engterm{Gumbel} law $\exp(-e^{-(x-\nu)/\beta})$, the limit for
light-tailed parents such as the exponential, normal, and gamma, with a
tail that decays exponentially; $\xi > 0$ is the \engterm{Fr\'echet}
law, the limit for heavy-tailed parents (Pareto, Student-$t$, Cauchy),
with a polynomial tail and, for $\xi \geq 1$, no finite mean; $\xi < 0$
is the \engterm{Weibull} (reversed) law, the limit for parents with a
finite upper endpoint such as the uniform and the beta. The shape
parameter is the single number that decides how the worst case scales,
and estimating it from data is the central task of extreme-value
engineering. The practical lesson runs opposite to the central-limit
instinct: the maximum is \emph{not} asymptotically Gaussian, it is
dominated by the heaviest tail of the parent, and a designer who
extrapolates a Gaussian fit of the body into the tail (the error
Figure~\ref{fig:vol02:ch13:fat-tail} draws and the 2008 case of section
13.9 lived) will misprice the rare large load by the gap between the
Gaussian tail and the true GEV tail. The companion discipline is the
\engterm{peaks-over-threshold} method, which fits a generalised Pareto
distribution to all exceedances above a high threshold rather than to
block maxima, using the data more efficiently; Volume X (failure and
reliability) develops both, and the $100$-year-load arithmetic above is
the order-statistic seed from which the whole theory grows.
\end{mastery}

\begin{mastery}
\textbf{Deriving a distribution with the moment generating function.}
The MGF turns a question about a sum into a question about a product,
and the cleanest demonstration derives the gamma from the
exponential. The MGF of $X \sim \mathrm{Exp}(\lambda)$ is
\[
M_{X}(t) = \int_{0}^{\infty} e^{tx} \lambda e^{-\lambda x} \dd x
= \frac{\lambda}{\lambda - t}, \qquad t < \lambda.
\]
For a sum $S = \sum_{i=1}^{\alpha} X_{i}$ of $\alpha$ independent
$\mathrm{Exp}(\lambda)$ variables, the MGF of the sum is the product
of the MGFs,
\[
M_{S}(t) = \left( \frac{\lambda}{\lambda - t} \right)^{\alpha},
\]
which is exactly the MGF of $\mathrm{Gamma}(\alpha, \lambda)$. The
uniqueness theorem for moment generating functions (when the MGF
exists in a neighbourhood of zero, it determines the distribution)
finishes the argument: the sum of $\alpha$ memoryless waits is gamma,
which is why the time to the $\alpha$th arrival of a Poisson process
is gamma and why an $\alpha$-stage service in a queue has a gamma
service time. Differentiating $M_{S}$ at $t = 0$ recovers the moments
without integrating the gamma density: $M_{S}'(0) = \alpha/\lambda$
gives the mean and $M_{S}''(0) = \alpha(\alpha+1)/\lambda^{2}$ gives
$\E[S^{2}]$, so $\Var(S) = \alpha/\lambda^{2}$, matching the catalogue
entry.
\end{mastery}

\subsection{Conditional expectation as a random variable}

The conditional expectation $\E[Y \mid X = x]$ computed in section 13.5
is a number, one for each value $x$. Collecting those numbers across the
random outcome of $X$ promotes the object to a random variable. We write
$\E[Y \mid X]$ for the random variable that takes the value $\E[Y \mid
X = x]$ on the event $\{X = x\}$. It is a function of $X$ alone, and it
is the best predictor of $Y$ from $X$ in the mean-square sense: among all
functions $g(X)$, the choice $g(X) = \E[Y \mid X]$ minimises
$\E[(Y - g(X))^{2}]$. The regression line of the bivariate-Gaussian
worked example is this object specialised to the jointly Gaussian case,
where the best predictor happens to be linear; for a general joint law
the best predictor is the conditional-mean curve, which need not be a
line.

Two identities make the random-variable view pay. The \engterm{tower
property} (the law of iterated expectation) states that averaging the
conditional expectation over $X$ recovers the unconditional mean,
\[
\E\!\big[\,\E[Y \mid X]\,\big] = \E[Y].
\]
The inner expectation removes the randomness in $Y$ given $X$; the outer
expectation removes the remaining randomness in $X$. The
\engterm{law of total variance} decomposes the spread of $Y$ into the
part explained by $X$ and the part left over,
\[
\Var(Y) = \underbrace{\E\!\big[\,\Var(Y \mid X)\,\big]}
_{\text{within-group (unexplained)}}
+ \underbrace{\Var\!\big(\,\E[Y \mid X]\,\big)}
_{\text{between-group (explained)}}.
\]
The first term is the average residual variance after conditioning; the
second is the variance of the conditional mean across the conditioning
variable. The decomposition is the probabilistic skeleton of the
analysis of variance in Volume I Chapter 8, of the variance-component
models in metrology, and of the bias-variance accounting that the
machine-learning chapters use.

\paragraph{Worked example: a random number of random parts (compound
sums).} A delivery contains a random number $N$ of parts, with $N \sim
\mathrm{Pois}(\lambda)$, and each part has an independent random mass
$M_{i}$ with mean $\mu$ and variance $\sigma^{2}$. The total shipped
mass is the \engterm{compound sum} $T = \sum_{i=1}^{N} M_{i}$, a sum with
a random number of terms. Conditioning on $N$ makes the inner problem an
ordinary fixed-length sum. By the tower property,
\[
\E[T] = \E\!\big[\,\E[T \mid N]\,\big] = \E[N\mu] = \lambda\mu,
\]
and by the law of total variance, using $\Var(T \mid N) = N\sigma^{2}$
and $\E[T \mid N] = N\mu$,
\[
\Var(T) = \E[N\sigma^{2}] + \Var(N\mu)
= \lambda\sigma^{2} + \mu^{2}\Var(N)
= \lambda\sigma^{2} + \mu^{2}\lambda
= \lambda(\sigma^{2} + \mu^{2}) = \lambda\,\E[M^{2}],
\]
where the Poisson identity $\Var(N) = \lambda$ collapses the two terms.
For $\lambda = 30$ parts per delivery, $\mu = 2\,\text{kg}$, and
$\sigma = 0.5\,\text{kg}$, the mean shipped mass is $60\,\text{kg}$ and
the variance is $30(0.25 + 4) = 127.5\,\text{kg}^{2}$, a standard
deviation of $11.3\,\text{kg}$. The two sources of variation, the random
part count and the random part mass, are separated cleanly by the
decomposition: the count contributes $\mu^{2}\lambda = 120$ and the
per-part mass contributes $\lambda\sigma^{2} = 7.5$, so the uncertainty
in the total is dominated by not knowing how many parts arrive, not by
the mass scatter within a known count. The same compound-sum arithmetic
prices aggregate insurance claims, total demand over a random number of
orders, and the total service time of a random number of jobs, which is
why the result is the entry point to the collective-risk models of
actuarial and operations engineering.

\begin{mastery}
\textbf{Conditional expectation, projection, and the orthogonality
principle.} In the space $L^{2}$ of square-integrable random variables
with inner product $\langle U, V \rangle = \E[UV]$, the conditional
expectation $\E[Y \mid X]$ is the orthogonal projection of $Y$ onto the
subspace of functions of $X$. Optimality is the projection theorem: the
residual $Y - \E[Y \mid X]$ is orthogonal to every function $g(X)$,
\[
\E\!\big[(Y - \E[Y \mid X])\,g(X)\big] = 0,
\]
which is the \engterm{orthogonality principle} that the Kalman filter,
least-squares regression, and Wiener filtering all instantiate. Taking
$g \equiv 1$ recovers the tower property; taking $g(X) = \E[Y \mid X]$
and expanding $\E[Y^{2}] = \E[(\E[Y\mid X])^{2}] + \E[(Y - \E[Y\mid
X])^{2}]$ recovers the law of total variance as the Pythagorean theorem
in $L^{2}$. The geometry is the reason a linear-Gaussian estimator is
both the best linear estimator and the best estimator outright: in the
Gaussian world the projection onto functions of $X$ and the projection
onto linear functions of $X$ coincide, so linearity costs nothing. When
the joint law is non-Gaussian the two projections differ, and the gap is
exactly the value a non-linear estimator can add over the best linear
one.
\end{mastery}

\subsection{Change of variables in several dimensions}

The single-variable change-of-variables formula extends to a smooth
invertible map $\vect{Y} = \vect{g}(\vect{X})$ on $\R^{n}$ with inverse
$\vect{h} = \vect{g}^{-1}$. The density transforms by the absolute value
of the determinant of the Jacobian of the inverse map,
\[
f_{\vect{Y}}(\vect{y}) = f_{\vect{X}}(\vect{h}(\vect{y}))\,
\bigg| \det \frac{\partial \vect{h}}{\partial \vect{y}} \bigg|,
\]
where $\partial\vect{h}/\partial\vect{y}$ is the matrix of partial
derivatives $\partial h_{i}/\partial y_{j}$. The determinant is the
local volume-scaling factor of the map, the multidimensional
generalisation of the $|h'(y)|$ stretch factor of the scalar case.
Volume II Chapter 11 derived the change-of-variables theorem for
multiple integrals; here it becomes the transformation law for joint
densities, and the discipline it forces is the same: identify the map,
invert it, compute the Jacobian determinant, and check that the
transformed density integrates to one over the image region.

\paragraph{Worked example: the Box-Muller transform.} Two independent
uniforms manufacture two independent standard normals, the identity
behind the Gaussian random-number generator the project relies on. Let
$U_{1}, U_{2} \sim \mathrm{Unif}(0, 1)$ independent, and define
\[
X = \sqrt{-2\ln U_{1}}\,\cos(2\pi U_{2}), \qquad
Y = \sqrt{-2\ln U_{1}}\,\sin(2\pi U_{2}).
\]
Reading the construction in polar form, $R = \sqrt{X^{2} + Y^{2}} =
\sqrt{-2\ln U_{1}}$ and $\Theta = 2\pi U_{2}$, so $R^{2} = -2\ln U_{1}$
is $\mathrm{Exp}(\tfrac12)$ by the inverse-transform identity of section
13.3, and $\Theta$ is uniform on $[0, 2\pi)$. The polar-to-Cartesian map
$(r, \theta) \mapsto (x, y) = (r\cos\theta, r\sin\theta)$ has Jacobian
determinant $r$, so the joint density of $(X, Y)$ is
\[
f_{X,Y}(x, y) = f_{R,\Theta}(r, \theta)\,\frac{1}{r}
= \underbrace{r\,e^{-r^{2}/2}}_{f_{R}(r)} \cdot
\underbrace{\frac{1}{2\pi}}_{f_{\Theta}(\theta)} \cdot \frac{1}{r}
= \frac{1}{2\pi}\,e^{-(x^{2}+y^{2})/2}
= \frac{1}{\sqrt{2\pi}}e^{-x^{2}/2}\cdot
\frac{1}{\sqrt{2\pi}}e^{-y^{2}/2},
\]
which factors as the product of two standard-normal densities. The
factor of $r$ from the polar density and the factor of $1/r$ from the
Jacobian cancel, which is the whole trick: the Rayleigh radius and the
uniform angle conspire to give an isotropic Gaussian. The derivation
earns its place because it is the cheapest correct way to generate
Gaussian samples (no normal-CDF inversion required), because the
cancellation is a clean instance of the multidimensional change-of-
variables, and because the same polar decomposition explains why the
chi-square with two degrees of freedom is exponential and why the
length and direction of an isotropic Gaussian vector are independent.

\input{volumes/02-form/ch13-probability/figures/fig-box-muller.tex}

\section{Common distributions: discrete and continuous}\pathtag{core}

Volume I Chapter 8 §8.2 introduced the normal, log-normal, Weibull,
Poisson, and exponential distributions at recognition level. We
expand the catalogue with the additional distributions an engineer
needs to recognise on sight, with their parameters, mechanism, and
moments \cite[ch.~4-5]{text:ross-probability}. The reader internalises
the table by use, not by memorisation; the exercises return to it
repeatedly.

\subsection{Discrete: Bernoulli, binomial, geometric, Poisson}

A \engterm{Bernoulli random variable} $X \sim \mathrm{Bern}(p)$
takes value $1$ with probability $p$ and $0$ with probability
$1 - p$. The mean is $p$ and the variance is $p(1 - p)$. The
Bernoulli is the indicator of a single yes/no event: a coin flip,
a single quality-control pass/fail, a single packet delivered or
dropped.

A \engterm{binomial random variable} $X \sim \mathrm{Bin}(n, p)$
counts the successes in $n$ independent Bernoulli trials with
success probability $p$,
\[
\Pr(X = k) = \binom{n}{k} p^{k} (1 - p)^{n - k},
\quad k = 0, 1, \ldots, n.
\]
The mean is $np$ and the variance is $np(1 - p)$. The binomial
arises whenever a fixed number of independent trials are summed,
each with the same success probability. Its independence assumption
is restrictive, and the failure section returns to it.

A \engterm{geometric random variable} $X \sim \mathrm{Geom}(p)$
counts the number of independent Bernoulli trials up to and
including the first success,
\[
\Pr(X = k) = (1 - p)^{k - 1} p, \quad k = 1, 2, \ldots
\]
The mean is $1/p$ and the variance is $(1 - p)/p^{2}$. The
geometric is the discrete analogue of the exponential and shares
its memoryless property: $\Pr(X > m + n \mid X > m) = \Pr(X > n)$.
The geometric models the trial count to first failure of a
constant-hazard system, the number of tries to first success in
sampling, and the wait time in discrete-time queueing.

\paragraph{Worked example: retransmissions on a lossy link.} A
packet is sent over a link that drops each transmission
independently with probability $q = 0.2$, and the sender retries
until an acknowledgement returns. The number of transmissions $X$
to first success is $\mathrm{Geom}(p)$ with $p = 1 - q = 0.8$. The
expected number of transmissions is $\E[X] = 1/p = 1.25$, so on
average one and a quarter sends deliver a packet. The probability
that more than three transmissions are needed is $\Pr(X > 3) =
(1 - p)^{3} = 0.2^{3} = 0.008$, eight in a thousand. If the link
budget allows at most two retries (three total transmissions), the
delivery probability per packet is $1 - 0.008 = 0.992$, and over a
$1000$-packet flow the expected number of undelivered packets is
about $8$. Doubling the retry budget to five total transmissions
drops the loss to $0.2^{5} = 3.2 \times 10^{-4}$ per packet. The
geometric tail is what a protocol designer trades against latency:
each extra retry multiplies the residual loss by $q$, a fixed factor
because the link is memoryless.

A \engterm{Poisson random variable} $X \sim \mathrm{Pois}(\lambda)$
takes values in $\{0, 1, 2, \ldots\}$ with
\[
\Pr(X = k) = \frac{\lambda^{k} e^{-\lambda}}{k!}.
\]
The mean and the variance both equal $\lambda$. The Poisson is the
limit of $\mathrm{Bin}(n, \lambda/n)$ as $n \to \infty$ with
$\lambda$ fixed; it models counts of rare independent events in a
fixed interval. Volume I Chapter 8 §8.2 gave the working examples.

\begin{mastery}
\textbf{Poisson as the limit of the binomial.} Fix $\lambda$ and
write $p = \lambda/n$ in the binomial mass function:
\[
\Pr(X = k) = \binom{n}{k}\left(\frac{\lambda}{n}\right)^{k}
\left(1 - \frac{\lambda}{n}\right)^{n - k}
= \frac{n!}{k!(n - k)! \, n^{k}} \lambda^{k}
\left(1 - \frac{\lambda}{n}\right)^{n}
\left(1 - \frac{\lambda}{n}\right)^{-k}.
\]
As $n \to \infty$, the falling factorial ratio
$n!/[(n - k)! \, n^{k}] \to 1$, the term $(1 - \lambda/n)^{n} \to
e^{-\lambda}$, and $(1 - \lambda/n)^{-k} \to 1$. What remains is
$\lambda^{k} e^{-\lambda}/k!$, the Poisson mass. The derivation is
the formal content of the rule of thumb that the binomial is well
approximated by the Poisson when $n$ is large and $p$ is small,
which is why counts of rare events (defects per wafer, calls per
minute, decays per second) are Poisson regardless of the precise
mechanism that generates them.
\end{mastery}

\paragraph{Worked example: acceptance sampling.} A lot is accepted
if a sample of $n = 20$ units contains at most one defective. The
true defect rate is $p = 0.05$. The number of defectives is
$X \sim \mathrm{Bin}(20, 0.05)$, and the acceptance probability is
\[
\Pr(X \leq 1) = (0.95)^{20} + 20(0.05)(0.95)^{19}
\approx 0.3585 + 0.3774 = 0.7358.
\]
A lot at the nominal $5\,\%$ defect rate is accepted about $74\,\%$
of the time by this plan. If the rate doubles to $p = 0.10$, the
acceptance probability falls to $(0.90)^{20} + 20(0.10)(0.90)^{19}
\approx 0.1216 + 0.2702 = 0.3918$. The plan's \engterm{operating
characteristic curve} is the acceptance probability as a function
of the true rate; the steepness of that curve between $p = 0.05$ and
$p = 0.10$ is the plan's discriminating power, and a buyer who wants
tighter discrimination must raise $n$ or lower the acceptance
number.

\subsection{Continuous: uniform, exponential, normal, gamma, beta}

A \engterm{uniform random variable} $X \sim \mathrm{Unif}(a, b)$ has
density $f(x) = 1/(b - a)$ for $a \leq x \leq b$ and zero elsewhere.
The mean is $(a + b)/2$ and the variance is $(b - a)^{2}/12$. The
uniform is the canonical maximum-entropy distribution on a bounded
interval and the first port of call when the engineer knows the
range of a quantity but nothing about its shape.

An \engterm{exponential random variable} $X \sim \mathrm{Exp}(\lambda)$
has density $f(x) = \lambda e^{-\lambda x}$ for $x \geq 0$, with
$\lambda > 0$ the rate parameter. The mean is $1/\lambda$ and the
variance is $1/\lambda^{2}$. The exponential is memoryless and
arises as the waiting time between events of a Poisson process with
rate $\lambda$. Volume I Chapter 8 §8.2 named the engineering
contexts.

\begin{mastery}
\textbf{Hazard rate and why constant hazard means exponential.} The
\engterm{hazard rate} of a positive random variable with density
$f$ and survival function $S(t) = \Pr(X > t)$ is
\[
h(t) = \frac{f(t)}{S(t)} = -\frac{\dd}{\dd t} \ln S(t),
\]
the instantaneous failure rate of a unit that has survived to time
$t$. Integrating, $S(t) = \exp(-\int_{0}^{t} h(u) \dd u)$, so the
survival function is determined by the hazard. A \emph{constant}
hazard $h(t) = \lambda$ gives $S(t) = e^{-\lambda t}$, which is the
exponential: constant hazard and the exponential law are the same
statement. The Weibull law $S(t) = \exp(-(\lambda t)^{k})$ has
hazard $h(t) = \lambda k (\lambda t)^{k-1}$, which rises with $t$
for shape $k > 1$ (wear-out) and falls for $k < 1$ (infant
mortality), and reduces to the exponential at $k = 1$. The bathtub
curve of reliability engineering is three hazard regimes stitched
together: a falling infant-mortality phase, a flat useful-life
phase where the exponential applies, and a rising wear-out phase.
Volume X develops the apparatus; the lesson here is that the
exponential is the right model only in the flat middle of the
bathtub.
\end{mastery}

A \engterm{normal random variable} $X \sim \mathcal{N}(\mu,
\sigma^{2})$ has density
\[
f(x) = \frac{1}{\sigma\sqrt{2\pi}}
\exp\!\left(-\frac{(x - \mu)^{2}}{2\sigma^{2}}\right).
\]
The mean is $\mu$, the variance is $\sigma^{2}$. The normal is the
limit distribution in the central limit theorem (section 13.6) and
the canonical model for measurement noise that aggregates many
small independent contributions. The standard normal has $\mu = 0$
and $\sigma = 1$, with CDF denoted $\Phi(z)$.

\paragraph{Worked example: tolerance yield under a normal model.}
A machined shaft has target diameter $25.000\,\text{mm}$ with a
two-sided tolerance of $\pm 0.050\,\text{mm}$. The process is
centred at $25.000\,\text{mm}$ with standard deviation $\sigma =
0.020\,\text{mm}$. The fraction of parts within tolerance is
\[
\Pr(24.950 < X < 25.050)
= \Phi\!\left(\frac{0.050}{0.020}\right)
- \Phi\!\left(\frac{-0.050}{0.020}\right)
= \Phi(2.5) - \Phi(-2.5) \approx 0.9876,
\]
so about $1.24\,\%$ of parts, or $12\,400$ per million, fall out of
tolerance. If the process drifts so the mean shifts to
$25.010\,\text{mm}$ while $\sigma$ holds, the upper limit is now
$(25.050 - 25.010)/0.020 = 2.0$ standard deviations away and the
lower limit is $(24.950 - 25.010)/0.020 = -3.0$, giving
$\Phi(2.0) - \Phi(-3.0) \approx 0.9772 - 0.0013 = 0.9759$: the
out-of-tolerance fraction more than doubles to about $2.4\,\%$ from
a drift of one quarter of the tolerance band. The asymmetry of the
loss under a centring error is the quantitative case for control
charts, and the calculation is the normal-tail arithmetic behind
the process-capability indices of Volume I Chapter 8.

A \engterm{gamma random variable} $X \sim \mathrm{Gamma}(\alpha,
\lambda)$ with shape $\alpha > 0$ and rate $\lambda > 0$ has
density
\[
f(x) = \frac{\lambda^{\alpha}}{\Gamma(\alpha)} x^{\alpha - 1}
e^{-\lambda x}, \quad x \geq 0,
\]
where $\Gamma(\alpha) = \int_{0}^{\infty} t^{\alpha - 1} e^{-t} \dd t$
is the gamma function. The mean is $\alpha/\lambda$ and the variance
is $\alpha/\lambda^{2}$. The gamma generalises the exponential
($\alpha = 1$) and the chi-square ($\alpha = k/2$, $\lambda = 1/2$).
A sum of $\alpha$ independent $\mathrm{Exp}(\lambda)$ variables is
$\mathrm{Gamma}(\alpha, \lambda)$, which makes the gamma a natural
model for the time to the $\alpha$th event of a Poisson process and
for service times in queueing systems where the service comprises
several memoryless stages.

A \engterm{beta random variable} $X \sim \mathrm{Beta}(\alpha,
\beta)$ with $\alpha, \beta > 0$ has density
\[
f(x) = \frac{x^{\alpha - 1}(1 - x)^{\beta - 1}}{B(\alpha, \beta)},
\quad 0 \leq x \leq 1,
\]
where $B(\alpha, \beta) = \Gamma(\alpha)\Gamma(\beta)/
\Gamma(\alpha + \beta)$ is the beta function. The mean is
$\alpha/(\alpha + \beta)$ and the variance is
$\alpha\beta/[(\alpha + \beta)^{2}(\alpha + \beta + 1)]$. The beta
is the canonical distribution on $[0, 1]$ and is the conjugate
prior for the Bernoulli/binomial likelihood: a $\mathrm{Beta}
(\alpha, \beta)$ prior, updated by $k$ successes in $n$ trials,
yields a $\mathrm{Beta}(\alpha + k, \beta + n - k)$ posterior. The
beta is therefore the tool of choice for proportion estimation in
Bayesian quality control.

\subsection{Recognition table}

The reader is expected to recognise each of these on sight and to
match it to a generative mechanism. Recognition is the working test:
when an engineer reads a problem statement, the distribution should
arrive without conscious selection. The exercises drill recognition;
the failure section in section 13.9 names the cost of fitting the
wrong family. Figure~\ref{fig:vol02:ch13:distribution-menagerie}
gives all nine on a single page. The accompanying
\repopath{volumes/02-form/ch13-probability/data/distributions_reference.csv}
carries the family, parameters, mean, variance, support, and
canonical mechanism in machine-readable form.

\input{volumes/02-form/ch13-probability/figures/fig-distribution-menagerie.tex}

\subsection{Choice of family: a working checklist}

A trained engineer takes the following six checks before committing
to a parametric family. The checklist is the recognition habit made
explicit:

\begin{itemize}[leftmargin=*]
  \item \textbf{Support.} Is the quantity discrete or continuous,
    bounded or unbounded, positive only or signed? A non-negative
    bounded quantity (a proportion, a percentage of capacity) is a
    beta or a truncated Gaussian, not a normal.
  \item \textbf{Mechanism.} Does the quantity arise as a count of
    rare events, a sum of independent components, a waiting time, a
    ratio, an extremum? Each mechanism points to its canonical
    family.
  \item \textbf{Skewness.} Is the empirical histogram symmetric,
    right-skewed, left-skewed? Right skew rules out the normal in
    favour of log-normal, Weibull, gamma; left skew is uncommon and
    often signals an upper bound.
  \item \textbf{Tail weight.} How heavy are the tails relative to a
    Gaussian? An empirical kurtosis above $3$ disqualifies the
    Gaussian for tail-risk questions even when the centre looks
    Gaussian.
  \item \textbf{Memorylessness.} Does the quantity have a
    constant-hazard interpretation (the conditional distribution of
    the remaining waiting time does not depend on the elapsed time)?
    If yes, exponential or geometric. If no, gamma, Weibull, or a
    non-parametric hazard.
  \item \textbf{Conjugacy.} If the distribution is a prior or a
    likelihood inside a Bayesian update loop, is a conjugate pair
    available (beta-binomial, gamma-Poisson, normal-normal)?
    Conjugacy buys computational tractability without giving up
    flexibility, provided the conjugate prior can express the actual
    prior knowledge.
\end{itemize}

The checklist does not guarantee the right family; no checklist does.
It rules out the wrong families fast enough that the engineer reaches
the candidate set the same way an experienced clinician reaches a
differential diagnosis: by recognition first, by ruling out second,
by goodness-of-fit testing last.

\section{Joint distributions, independence, correlation}\pathtag{core}

A vector of random variables on a common probability space has a
joint distribution. The joint distribution carries strictly more
information than the marginals; recovering the joint from the
marginals requires an additional assumption (independence is one;
specifying the copula is the general one).

\subsection{Joint, marginal, and conditional distributions}

For two random variables $X, Y$ on the same space, the
\engterm{joint CDF} is
\[
F_{X, Y}(x, y) = \Pr(X \leq x, Y \leq y).
\]
The \engterm{joint PMF} (in the discrete case) is $p_{X, Y}(x_{i},
y_{j}) = \Pr(X = x_{i}, Y = y_{j})$. The \engterm{joint PDF} (in
the continuous case) is the function $f_{X, Y}$ with
\[
\Pr((X, Y) \in A) = \iint_{A} f_{X, Y}(x, y) \,\dd x \,\dd y
\]
for measurable $A \subseteq \R^{2}$.

The \engterm{marginal} distribution of $X$ is recovered by summing
or integrating out $Y$:
\[
p_{X}(x_{i}) = \sum_{j} p_{X, Y}(x_{i}, y_{j})
\quad\text{or}\quad
f_{X}(x) = \int_{-\infty}^{\infty} f_{X, Y}(x, y) \,\dd y.
\]
The marginals do not determine the joint. Two random vectors with
identical marginals can have very different joint structures, and
the difference is exactly the engineering content of dependence.

The \engterm{conditional density} of $Y$ given $X = x$, when
$f_{X}(x) > 0$, is
\[
f_{Y \mid X}(y \mid x) = \frac{f_{X, Y}(x, y)}{f_{X}(x)}.
\]
Conditioning on a continuous random variable is delicate (the event
$\{X = x\}$ has probability zero), but the formula above gives the
working object.

\paragraph{Worked example: a joint PMF for two quality grades.}
A batch is graded by two inspectors, each assigning a unit to grade
$1$, $2$, or $3$. The joint mass function of $(X, Y)$, the two
grades for a random unit, is the table
\[
\begin{array}{c|ccc|c}
 & Y=1 & Y=2 & Y=3 & p_X \\ \hline
X=1 & 0.30 & 0.08 & 0.02 & 0.40 \\
X=2 & 0.06 & 0.24 & 0.10 & 0.40 \\
X=3 & 0.01 & 0.05 & 0.14 & 0.20 \\ \hline
p_Y & 0.37 & 0.37 & 0.26 & 1.00
\end{array}
\]
The right margin and bottom margin are the marginals, obtained by
summing each row and each column. The conditional distribution of
the second inspector's grade given the first assigned grade $2$ is
the middle row renormalised by $p_{X}(2) = 0.40$:
$\Pr(Y = 1 \mid X = 2) = 0.06/0.40 = 0.15$,
$\Pr(Y = 2 \mid X = 2) = 0.24/0.40 = 0.60$,
$\Pr(Y = 3 \mid X = 2) = 0.10/0.40 = 0.25$. The two inspectors
agree ($X = Y$) with probability $0.30 + 0.24 + 0.14 = 0.68$. They
are not independent: under independence $\Pr(X = 1, Y = 1)$ would be
$p_{X}(1) p_{Y}(1) = 0.40 \cdot 0.37 = 0.148$, far below the
observed $0.30$, which is the diagonal concentration that any
agreement-between-raters statistic (Cohen's $\kappa$) is built to
quantify.

\subsection{Independence of random variables}

Random variables $X, Y$ are \engterm{independent} when
$F_{X, Y}(x, y) = F_{X}(x) F_{Y}(y)$ for every $x, y$, equivalently
when $f_{X, Y} = f_{X} f_{Y}$ in the continuous case or
$p_{X, Y} = p_{X} p_{Y}$ in the discrete case. Independence is the
condition under which expectations factor: for independent $X, Y$
with finite means,
\[
\E[g(X) h(Y)] = \E[g(X)] \E[h(Y)]
\]
for any functions $g, h$ for which the expectations exist. The
factorisation is the workhorse identity of probability, and its
absence is where modelling work concentrates.

\subsection{Covariance and correlation}

The \engterm{covariance} of $X$ and $Y$ is
\[
\Cov(X, Y) = \E[(X - \E X)(Y - \E Y)] = \E[XY] - \E[X]\E[Y].
\]
The \engterm{correlation coefficient} is
\[
\Cor(X, Y) = \frac{\Cov(X, Y)}{\sigma_{X} \sigma_{Y}}.
\]
Correlation is unitless and lies in $[-1, 1]$. The bound follows
from the Cauchy-Schwarz inequality applied to $(X - \E X, Y - \E Y)$
in the inner-product space of square-integrable random variables.

The covariance matrix of a random vector $\vect{X} = (X_{1}, \ldots,
X_{n})\transpose$ is the $n \times n$ matrix $\Sigma$ with entries
$\Sigma_{ij} = \Cov(X_{i}, X_{j})$. The matrix is symmetric and
positive semidefinite. Its diagonal entries are the variances; its
off-diagonal entries carry every pairwise dependence the engineer
has acknowledged. Volume II Chapter 9 (linear algebra) gave the
spectral apparatus we now need: $\Sigma = Q \Lambda Q\transpose$ for
some orthogonal $Q$ and non-negative diagonal $\Lambda$, which is
the basis of principal-component analysis as a dimensionality
reduction tool.

\paragraph{Worked example: the variance of a sum of correlated
loads.} Two structural members each carry a random load with mean
$\mu = 10\,\text{kN}$ and standard deviation $\sigma = 2\,\text{kN}$.
The combined load on the shared support is $S = X_{1} + X_{2}$. The
mean is $20\,\text{kN}$ regardless of dependence, by linearity. The
standard deviation depends on the correlation $\rho$:
\[
\Var(S) = \sigma^{2} + \sigma^{2} + 2\rho \sigma^{2}
= 2\sigma^{2}(1 + \rho),
\qquad \sigma_{S} = \sigma\sqrt{2(1 + \rho)}.
\]
At $\rho = 0$ (independent loads), $\sigma_{S} = 2\sqrt{2} \approx
2.83\,\text{kN}$. At $\rho = 1$ (perfectly correlated, the two loads
rise and fall together), $\sigma_{S} = 2 \cdot 2 = 4\,\text{kN}$, a
factor of $\sqrt{2}$ larger. At $\rho = -1$ (perfectly
anti-correlated), $\sigma_{S} = 0$: the loads cancel and the support
sees a deterministic $20\,\text{kN}$. A designer who assumes
independent loads and sizes the support to a $3\sigma$ envelope of
$20 + 3(2.83) = 28.5\,\text{kN}$ has under-budgeted the margin if the
loads are in fact correlated through a common cause such as a shared
gust or a shared occupancy pattern, where the true $3\sigma$ figure
is $20 + 3(4) = 32\,\text{kN}$. The diversification benefit that
makes a portfolio of independent risks safer than its components is
the same identity read with $\rho < 1$, and its disappearance when
$\rho \to 1$ is exactly the failure pattern of section 13.9.

\subsection{Independence vs zero correlation}

Two random variables that are independent are uncorrelated: the
expectation factors and the covariance vanishes. The converse is
false. Zero correlation does not imply independence, except for the
joint Gaussian distribution where the converse does hold.

The standard counterexample is $X \sim \mathcal{N}(0, 1)$ and
$Y = X^{2}$. Then $\Cov(X, Y) = \E[X \cdot X^{2}] - \E[X] \E[X^{2}]
= \E[X^{3}] - 0 \cdot 1 = 0$. The correlation is zero, yet $Y$ is a
deterministic function of $X$; knowing $X$ pins down $Y$ exactly.
The engineering consequence is that a Pearson correlation
coefficient near zero does not certify independence. A scatter plot
or a non-linear dependence measure (mutual information, distance
correlation) is required when the question is independence rather
than the absence of a linear trend. The three panels of
Figure~\ref{fig:vol02:ch13:independence-correlation} show the
distinction directly: an isotropic Gaussian cloud (independent),
a U-shape with $\rho = 0$ (dependent and uncorrelated), and a
tilted Gaussian cloud at $\rho = 0.85$.

\input{volumes/02-form/ch13-probability/figures/fig-independence-correlation.tex}

\subsection{The bivariate Gaussian}

A pair $(X, Y)$ is \engterm{jointly Gaussian} when every linear
combination $aX + bY$ is Gaussian. The joint density, when the
covariance matrix $\Sigma$ is non-singular, is
\[
f_{X, Y}(x, y) = \frac{1}{2\pi |\Sigma|^{1/2}}
\exp\!\left(-\frac{1}{2} \vect{z}\transpose \Sigma^{-1} \vect{z}\right),
\]
where $\vect{z} = (x - \mu_{X}, y - \mu_{Y})\transpose$. The
contours of the density are ellipses aligned with the eigenvectors
of $\Sigma$. For the joint Gaussian and only for the joint
Gaussian, zero correlation implies independence. The reader who
treats every elliptical scatter plot as Gaussian without checking
gives up the converse direction the moment the data are heavy-tailed
or asymmetric.

\paragraph{Worked example: predicting one measurement from another.}
Two co-located sensors measure the same temperature with jointly
Gaussian readings: $X$ has mean $20\,\degC$ and standard deviation
$\sigma_{X} = 0.5\,\degC$; $Y$ has mean $20\,\degC$ and standard
deviation $\sigma_{Y} = 0.8\,\degC$; their correlation is
$\rho = 0.9$. The first sensor reads $X = 21.0\,\degC$. The best
prediction of the second sensor's reading is the conditional mean,
\[
\E[Y \mid X = 21.0] = \mu_{Y} + \rho \frac{\sigma_{Y}}{\sigma_{X}}
(x - \mu_{X}) = 20 + 0.9 \cdot \frac{0.8}{0.5} \cdot (21.0 - 20)
= 20 + 1.44 = 21.44\,\degC,
\]
and the residual uncertainty is the conditional standard deviation,
\[
\sigma_{Y \mid X} = \sigma_{Y}\sqrt{1 - \rho^{2}}
= 0.8 \sqrt{1 - 0.81} = 0.8 \cdot 0.436 \approx 0.349\,\degC.
\]
Conditioning on the first sensor cuts the second sensor's
uncertainty from $0.8$ to $0.35\,\degC$, a factor of
$\sqrt{1 - \rho^{2}}$. The conditional mean as a function of $x$ is
the regression line, and the conditional variance is the residual
variance left after the linear prediction. This is the calculation a
Kalman filter performs at every step, and the derivation behind it
is the subject of the bivariate-Gaussian derivation exercise.

\subsection{A continuous joint density, end to end}

The discrete table above kept the bookkeeping visible. The continuous
case replaces sums by integrals and asks the engineer to track the
support, which is where most errors enter. Take the joint density
\[
f_{X, Y}(x, y) = c\,(x + y), \qquad 0 \leq x \leq 1,\ 0 \leq y \leq 1,
\]
and zero elsewhere, a stylised model for two wear indices that both
rise toward end of life. The normalising constant comes from the
requirement that the density integrate to one:
\[
\int_{0}^{1}\!\!\int_{0}^{1} c(x + y)\,\dd x\,\dd y
= c \int_{0}^{1}\left( \tfrac12 + y \right)\dd y
= c \left( \tfrac12 + \tfrac12 \right) = c,
\]
so $c = 1$. The marginal of $X$ integrates out $y$,
\[
f_{X}(x) = \int_{0}^{1} (x + y)\,\dd y = x + \tfrac12,
\qquad 0 \leq x \leq 1,
\]
which is not uniform: larger $x$ is more likely because the joint
mass leans toward the upper-right corner. By symmetry $f_{Y}(y) = y +
\tfrac12$. The product of the marginals is $(x + \tfrac12)(y +
\tfrac12) = xy + \tfrac12 x + \tfrac12 y + \tfrac14$, which does not
equal the joint $x + y$, so $X$ and $Y$ are dependent. The conditional
density of $Y$ given $X = x$ is the joint over the marginal,
\[
f_{Y \mid X}(y \mid x) = \frac{x + y}{x + \tfrac12},
\qquad 0 \leq y \leq 1,
\]
a density on $[0, 1]$ that tilts more steeply toward large $y$ when
$x$ is small. The conditional expectation follows by integration,
\[
\E[Y \mid X = x] = \int_{0}^{1} y\,\frac{x + y}{x + \tfrac12}\,\dd y
= \frac{\tfrac12 x + \tfrac13}{x + \tfrac12}
= \frac{3x + 2}{6x + 3},
\]
which decreases from $\tfrac23$ at $x = 0$ to $\tfrac59$ at $x = 1$:
observing a high first index lowers the expected second index slightly,
the negative-dependence signature of this density. A probability over
a region integrates the joint directly; for the event that both
indices stay below one half,
\[
\Pr\!\left(X \leq \tfrac12,\, Y \leq \tfrac12\right)
= \int_{0}^{1/2}\!\!\int_{0}^{1/2} (x + y)\,\dd x\,\dd y
= 2 \cdot \tfrac12 \cdot \tfrac18 = \tfrac18,
\]
against the $\tfrac14 \cdot \tfrac14 = \tfrac1{16}$ an engineer would
get by multiplying the two marginal probabilities $\Pr(X \leq
\tfrac12) = \int_{0}^{1/2}(x + \tfrac12)\dd x = \tfrac38$. The point
of the worked integral is the discipline it forces: every continuous
joint problem is a careful integral over the correct support, and the
independence shortcut is available only after the product check above
has passed.

\begin{mastery}
\textbf{The multivariate normal.} The bivariate Gaussian generalises
to $n$ dimensions. A random vector $\vect{X} = (X_{1}, \ldots,
X_{n})\transpose$ is \engterm{multivariate normal} with mean
$\vect{\mu}$ and covariance $\Sigma$ (symmetric positive definite)
when its density is
\[
f_{\vect{X}}(\vect{x}) = \frac{1}{(2\pi)^{n/2} |\Sigma|^{1/2}}
\exp\!\left( -\tfrac12 (\vect{x} - \vect{\mu})\transpose
\Sigma^{-1} (\vect{x} - \vect{\mu}) \right).
\]
Three properties make it the workhorse of estimation and control.
\emph{Closure under linear maps}: if $\vect{X}$ is multivariate
normal then $\mat{A}\vect{X} + \vect{b}$ is multivariate normal with
mean $\mat{A}\vect{\mu} + \vect{b}$ and covariance $\mat{A}\Sigma
\mat{A}\transpose$, which is why a Kalman filter stays Gaussian
through every linear update. \emph{Marginals and conditionals are
Gaussian}: partition $\vect{X} = (\vect{X}_{1}, \vect{X}_{2})$ with
the matching block covariance; the conditional $\vect{X}_{1} \mid
\vect{X}_{2} = \vect{x}_{2}$ is Gaussian with mean $\vect{\mu}_{1} +
\Sigma_{12}\Sigma_{22}^{-1}(\vect{x}_{2} - \vect{\mu}_{2})$ and
covariance $\Sigma_{11} - \Sigma_{12}\Sigma_{22}^{-1}\Sigma_{21}$,
the Schur complement that is the engine of Gaussian-process
regression. \emph{Zero covariance is independence}: for the
multivariate normal, and only here in general, a block-diagonal
$\Sigma$ means the blocks are independent. The eigendecomposition
$\Sigma = Q\Lambda Q\transpose$ rotates the correlated coordinates
into independent ones, the whitening transform that underlies
principal-component analysis and the Mahalanobis distance
$(\vect{x} - \vect{\mu})\transpose\Sigma^{-1}(\vect{x} - \vect{\mu})$,
which is the squared distance in standard-deviation units and is
$\chi^{2}_{n}$-distributed under the model.
\end{mastery}

\begin{mastery}
\textbf{Copulas: separating margins from dependence.} Sklar's theorem
states that every joint CDF $F_{X, Y}$ factors as $F_{X, Y}(x, y) =
C(F_{X}(x), F_{Y}(y))$ for a \engterm{copula} $C$, a joint CDF on the
unit square with uniform margins, and $C$ is unique where the margins
are continuous. The decomposition separates two modelling decisions
the marginals and the dependence structure, which an engineer can
calibrate from different data. The independence copula is $C(u, v) =
uv$; the comonotone (perfect-dependence) copula is $\min(u, v)$. The
Gaussian copula, $C(u, v) = \Phi_{\rho}(\Phi^{-1}(u), \Phi^{-1}(v))$,
takes its dependence from a bivariate normal but accepts arbitrary
margins. Its defining limitation is asymptotic tail independence: for
$\rho < 1$ the probability that both variables are jointly extreme
decays faster than either margin alone, so the Gaussian copula assigns
near-zero probability to simultaneous tail events. That property is
exactly what made the Gaussian-copula default model of section 13.9
understate joint default in the 2008 crisis; a $t$-copula or a
Clayton copula, which carry positive lower-tail dependence, would have
priced the cluster risk the Gaussian copula erased. The copula is the
formal home of the lesson that identical marginals can hide radically
different joint-tail behaviour.
\end{mastery}

\subsection{Conditioning in three dimensions, worked}

The Schur-complement formula of the multivariate-normal mastery box is
abstract until run on numbers. Three co-located strain gauges on a beam
report jointly Gaussian readings $\vect{X} = (X_{1}, X_{2}, X_{3})$ with
common mean $\vect{\mu} = (100, 100, 100)\,\mu\varepsilon$ (microstrain)
and covariance
\[
\Sigma = \begin{pmatrix}
4 & 3 & 1 \\
3 & 9 & 2 \\
1 & 2 & 16
\end{pmatrix}\,\mu\varepsilon^{2}.
\]
Gauges $1$ and $2$ are well correlated; gauge $3$ is noisier and only
weakly coupled. Two of the gauges report and we predict the third. Take
the partition $\vect{X}_{1} = X_{3}$ (the target) and $\vect{X}_{2} =
(X_{1}, X_{2})$ (the observed pair), with the corresponding blocks
\[
\Sigma_{11} = 16, \quad
\Sigma_{12} = (1, 2), \quad
\Sigma_{22} = \begin{pmatrix} 4 & 3 \\ 3 & 9 \end{pmatrix}.
\]
The observed pair reads $(X_{1}, X_{2}) = (104, 98)$, a deviation
$\vect{x}_{2} - \vect{\mu}_{2} = (4, -2)$ from the mean. The inverse of
the $2 \times 2$ observed block is
\[
\Sigma_{22}^{-1} = \frac{1}{4 \cdot 9 - 3^{2}}
\begin{pmatrix} 9 & -3 \\ -3 & 4 \end{pmatrix}
= \frac{1}{27}\begin{pmatrix} 9 & -3 \\ -3 & 4 \end{pmatrix}.
\]
The gain vector is $\Sigma_{12}\Sigma_{22}^{-1} = (1, 2) \cdot
\tfrac{1}{27}\begin{psmallmatrix} 9 & -3 \\ -3 & 4 \end{psmallmatrix}
= \tfrac{1}{27}(9 - 6,\ -3 + 8) = \tfrac{1}{27}(3, 5)$. The conditional
mean of the third gauge is then
\[
\E[X_{3} \mid X_{1} = 104, X_{2} = 98]
= 100 + \tfrac{1}{27}(3, 5)\binom{4}{-2}
= 100 + \tfrac{1}{27}(12 - 10) = 100 + 0.074 = 100.07\,\mu\varepsilon,
\]
barely moved from the prior mean because gauge $3$ is weakly coupled to
the pair. The conditional variance is the Schur complement,
\[
\Var(X_{3} \mid X_{1}, X_{2}) = \Sigma_{11} -
\Sigma_{12}\Sigma_{22}^{-1}\Sigma_{21}
= 16 - \tfrac{1}{27}(3, 5)\binom{1}{2}
= 16 - \tfrac{13}{27} = 16 - 0.481 = 15.52\,\mu\varepsilon^{2},
\]
so observing two gauges trims the third gauge's standard deviation only
from $4.00$ to $3.94\,\mu\varepsilon$. The numbers carry the lesson that
the value of an extra sensor depends entirely on its correlation with
the ones already read: a tightly coupled gauge would collapse the
conditional variance, while gauge $3$ here is nearly redundant as a
predictor and nearly fresh as an independent measurement. The same
arithmetic, applied recursively as each observation arrives, is the
update step of the Kalman filter of Volume IX, and the conditional
variance shrinking by the Schur complement is the filter's covariance
update made concrete on three numbers.

\begin{mastery}
\textbf{Cumulants and the exponential family.} The logarithm of the
moment generating function, $K_{X}(t) = \ln M_{X}(t)$, is the
\engterm{cumulant generating function}; its Taylor coefficients are the
\engterm{cumulants} $\kappa_{1}, \kappa_{2}, \ldots$, with $\kappa_{1} =
\mu$, $\kappa_{2} = \sigma^{2}$, $\kappa_{3}$ the third central moment
(unnormalised skewness numerator), and $\kappa_{4}$ the excess-kurtosis
numerator. Cumulants beat moments for sums because the cumulant
generating function of a sum of independents is the \emph{sum} of the
cumulant generating functions, so cumulants add: $\kappa_{r}(X + Y) =
\kappa_{r}(X) + \kappa_{r}(Y)$ for independent $X, Y$ at every order.
The Gaussian is the unique law whose cumulants vanish beyond the second,
which is the cleanest statement of what ``Gaussian'' means and why the
central limit theorem standardises away every higher cumulant: under the
$\sqrt{n}$ scaling, $\kappa_{r}$ of the standardised sum decays like
$n^{1 - r/2}$, so all orders above the second wash out. A wide swath of
the named distributions of section 13.4 belong to the
\engterm{exponential family}, densities of the form $f(x \mid \theta) =
h(x)\exp(\eta(\theta)\transpose T(x) - A(\theta))$ with $T(x)$ the
\engterm{sufficient statistic} and $A(\theta)$ the
\engterm{log-partition function} whose derivatives generate the cumulants
of $T$. The Bernoulli, binomial, Poisson, exponential, gamma, beta, and
Gaussian are all exponential-family members; the family is the reason
conjugate priors exist (the conjugate prior is itself exponential-family
with $A$ as its kernel), the reason the maximum-likelihood equations
reduce to matching the observed sufficient statistic to its expectation,
and the unifying structure behind generalised linear models. The Cauchy
and the Student-$t$ are not exponential-family, which is the formal
counterpart of their refusal to admit a moment generating function and
their heavy tails.
\end{mastery}

\section{Limit theorems: LLN and CLT}\pathtag{core}

Two theorems carry the connection between probability theory and
statistical inference. The Law of Large Numbers (LLN) says that
sample averages converge to the population mean. The Central Limit
Theorem (CLT) says that sample averages are approximately Gaussian
around the population mean, with a known scale. Volume II Chapter
14 takes both theorems forward into formal inference; Volume IX
takes them forward into queueing and Volume X into reliability.

\subsection{Modes of convergence}

Convergence of random variables admits more than one mode. We name
two; the others are listed in any introductory text
\cite[ch.~5]{text:wasserman2004}.

A sequence $X_{n}$ \engterm{converges in probability} to $X$, written
$X_{n} \to_{P} X$, when for every $\varepsilon > 0$,
\[
\Pr(|X_{n} - X| > \varepsilon) \to 0 \quad\text{as } n \to \infty.
\]
A sequence $X_{n}$ \engterm{converges in distribution} to $X$,
written $X_{n} \to_{D} X$, when $F_{X_{n}}(x) \to F_{X}(x)$ at every
continuity point $x$ of $F_{X}$.

Convergence in probability implies convergence in distribution; the
converse holds when the limit is a constant. Convergence almost
surely is stronger than convergence in probability and is what the
strong LLN delivers. The reader needs the names; the working
content sits in the two theorems below.

\subsection{The Law of Large Numbers}

Let $X_{1}, X_{2}, \ldots$ be independent and identically
distributed (iid) random variables with finite mean
$\mu = \E[X_{i}]$. The \engterm{sample mean} is
$\bar X_{n} = (1/n) \sum_{i=1}^{n} X_{i}$.

The \engterm{Weak Law of Large Numbers} states $\bar X_{n} \to_{P}
\mu$. The \engterm{Strong Law of Large Numbers} states $\bar X_{n}
\to \mu$ almost surely: $\Pr(\lim_{n \to \infty} \bar X_{n} = \mu)
= 1$ \cite[ch.~8]{text:ross-probability}.

The LLN is the formal statement that frequencies converge to
probabilities and sample averages converge to expectations. It is
the licence the engineer has to use a sample mean as an estimate of
a population mean, the empirical defect rate as an estimate of the
true defect rate, and a Monte Carlo average as an estimate of an
integral. The convergence is at the rate $1/\sqrt{n}$ in standard
deviation: the sample mean's spread shrinks as the square root of
the sample size, which is the $1/\sqrt{n}$ wall the chapter on
engineering statistics named.

The LLN requires three conditions and breaks when any of them
fails. First, the mean must exist: a Cauchy sample average does not
converge, no matter how large $n$ is. Second, the variables must be
identically distributed: a non-stationary process can have a sample
average that wanders. Third, the variables must be independent (or
satisfy a weaker mixing condition): heavily correlated samples have
an effective sample size much smaller than $n$, and the convergence
slows or fails entirely.

\paragraph{Worked example: Monte Carlo integration and its rate.}
The LLN licenses estimating an integral as a sample average. To
compute $\theta = \int_{0}^{1} e^{-x^{2}} \dd x$, draw $U_{i} \sim
\mathrm{Unif}(0, 1)$ and average $g(U_{i}) = e^{-U_{i}^{2}}$. The
estimator $\hat\theta_{n} = (1/n) \sum_{i} e^{-U_{i}^{2}}$ converges
to $\theta \approx 0.7468$ by the LLN. The variance of a single
term is $\Var(g(U)) = \E[g^{2}] - \theta^{2} = \int_{0}^{1}
e^{-2x^{2}} \dd x - 0.7468^{2} \approx 0.5981 - 0.5577 = 0.0404$,
so the estimator's standard error is $\sqrt{0.0404/n} \approx
0.201/\sqrt{n}$. Reaching three-digit accuracy (standard error
$10^{-3}$) requires $n \approx (0.201/10^{-3})^{2} \approx
40\,000$ samples. The defining feature is the $1/\sqrt{n}$ rate:
each additional digit of accuracy costs a hundredfold increase in
samples, independent of the dimension of the integral. That
dimension-independence is why Monte Carlo beats grid quadrature for
high-dimensional integrals, where a grid of $m$ points per axis in
$d$ dimensions needs $m^{d}$ evaluations while Monte Carlo needs a
number set only by the target accuracy and the integrand's
variance.

\subsection{The Central Limit Theorem}

Let $X_{1}, X_{2}, \ldots$ be iid with finite mean $\mu$ and finite
variance $\sigma^{2}$. The \engterm{Central Limit Theorem} states
that
\[
\frac{\bar X_{n} - \mu}{\sigma / \sqrt{n}} \to_{D} \mathcal{N}(0, 1)
\quad\text{as } n \to \infty.
\]
Equivalently, $\sqrt{n}(\bar X_{n} - \mu) \to_{D} \mathcal{N}(0,
\sigma^{2})$, and for large $n$,
\[
\bar X_{n} \approx \mathcal{N}(\mu, \sigma^{2}/n).
\]
The CLT supplies the distributional shape that the LLN does not.
Whatever the distribution of the $X_{i}$, as long as the variance
is finite, the standardised sample mean is approximately Gaussian.
This is the formal apparatus behind the Gaussian assumption that
underwrites confidence intervals, hypothesis tests, and standard
errors throughout engineering practice.

The CLT is asymptotic. The rate at which the Gaussian
approximation becomes good depends on the third moment of the
underlying distribution; the Berry-Esseen theorem gives a uniform
bound \cite[ch.~5]{text:wasserman2004}. For nearly symmetric
distributions, $n = 30$ is often adequate; for heavily skewed
distributions, $n$ may need to be in the hundreds. For
heavy-tailed distributions with infinite variance, the CLT does not
apply at all; the limit, when it exists, is a non-Gaussian stable
law and the relevant convergence is at a rate slower than $1/\sqrt{n}$.

The CLT requires finite variance. A Pareto distribution with shape
parameter below $2$ has infinite variance, and the sample mean has a
non-Gaussian limit law. Engineers who fit a Gaussian to a
heavy-tailed quantity and then estimate confidence intervals via
$\sigma/\sqrt{n}$ understate the tail risk by an amount that
depends on the distribution; section 13.9 returns to this pattern.
Figure~\ref{fig:vol02:ch13:clt-convergence} shows the CLT for the
$\mathrm{Exp}(1)$ underlying distribution, where the standardised
sample mean is visibly Gaussian by $n = 30$ and indistinguishable
from a standard normal at $n = 100$. The companion script
\repopath{volumes/02-form/ch13-probability/code/clt_demo.py}
generates the histograms behind the figure for any sample size and
trial count of the reader's choice.

\input{volumes/02-form/ch13-probability/figures/fig-clt-convergence.tex}

\paragraph{Worked demonstration: the sum of dice.} A single fair
die has mean $\mu = 3.5$ and variance $\sigma^{2} = 35/12 \approx
2.917$. The sum of $n = 10$ independent dice has mean $35$ and
variance $350/12 \approx 29.17$, standard deviation $\approx 5.40$.
By the CLT the sum is approximately $\mathcal{N}(35, 29.17)$. The
probability the sum exceeds $45$ is then
\[
\Pr(S_{10} > 45) \approx 1 - \Phi\!\left(\frac{45 - 35}{5.40}\right)
= 1 - \Phi(1.85) \approx 0.032.
\]
Applying a continuity correction, $\Pr(S_{10} \geq 46) \approx 1 -
\Phi((45.5 - 35)/5.40) = 1 - \Phi(1.94) \approx 0.026$, which
matches the exact convolution result $0.0263$ to three figures even
though the underlying single-die distribution is discrete and
uniform, not Gaussian. Ten dice already deliver a Gaussian-quality
approximation because the uniform is symmetric and light-tailed.
The same arithmetic underwrites the rule that the sum of a dozen
independent rounding errors, each uniform on a half-bit interval,
is well modelled as Gaussian, which is why accumulated round-off in
a long calculation is treated as Gaussian noise rather than tracked
term by term.

\begin{mastery}
\textbf{Why the limit is Gaussian.} The characteristic function of
the standardised sum $Z_{n} = \sqrt{n}(\bar X_{n} - \mu)/\sigma$
factors over the independent terms. For a single standardised term
$W = (X - \mu)/\sigma$ with $\E[W] = 0$ and $\E[W^{2}] = 1$, the
characteristic function expands as $\E[e^{itW/\sqrt{n}}] = 1 -
t^{2}/(2n) + o(1/n)$. The product over $n$ terms is
\[
\E[e^{itZ_{n}}] = \left(1 - \frac{t^{2}}{2n} + o(1/n)\right)^{n}
\longrightarrow e^{-t^{2}/2},
\]
which is the characteristic function of $\mathcal{N}(0, 1)$. The
Gaussian appears because it is the unique fixed point of the
operation ``standardise and add'': the only finite-variance law
whose shape survives repeated convolution-and-rescaling. The $t^{2}$
term that drives the limit is the variance; the third-moment term,
which the expansion discards at order $1/n$, controls the
\emph{rate} of convergence and is the quantity the Berry-Esseen
bound makes explicit. When the variance is infinite, the quadratic
term diverges, the fixed point is a non-Gaussian stable law, and the
classical CLT does not hold.
\end{mastery}

\begin{estimation}
\textbf{Question.} A manufacturing process produces parts whose
weight $X$ has mean $\mu = 100\,\text{g}$ and standard deviation
$\sigma = 5\,\text{g}$. A box contains $n = 100$ parts. Estimate the
probability that the total box weight exceeds $10\,050\,\text{g}$.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The total weight is $S = \sum_{i=1}^{100} X_{i}$ with mean $n\mu =
10\,000\,\text{g}$ and standard deviation $\sigma\sqrt{n} = 50\,
\text{g}$. The threshold $10\,050\,\text{g}$ is one standard
deviation above the mean. The CLT, which is comfortable here for
$n = 100$, gives
\[
\Pr(S > 10\,050) \approx 1 - \Phi(1) \approx 0.159.
\]
A Chebyshev bound gives only $\Pr(|S - 10\,000| \geq 50) \leq 1$, an
empty bound at $k = 1$. The CLT is the right tool when the
distribution conditions are met; Chebyshev is the right tool when
they are not. The estimate the engineer takes forward is that
distributional information buys order-of-magnitude tighter
calculations, and the CLT is what makes that information
operationally available.
\end{estimation}

\begin{principle}[The square-root-of-$n$ wall]
The standard error of a sample mean of $n$ independent observations
shrinks as $\sigma/\sqrt{n}$. Halving the uncertainty requires
quadrupling the sample. This $1/\sqrt{n}$ rate is the same wall met
by Monte Carlo integration, by averaging to suppress measurement
noise, and by the running occupancy of an ergodic Markov chain. It
sets the price of precision across the book: precision is bought in
units of $\sqrt{n}$, and the engineer who needs one more decimal
digit pays a hundredfold in samples unless the variance itself can
be reduced (by a better estimator, stratification, or control
variates) or unless the rate can be beaten (by quasi-Monte Carlo,
which trades randomness for low-discrepancy sequences). The wall is
a property of independent averaging, not of the particular quantity
being averaged.
\end{principle}

\begin{mastery}
\textbf{Concentration: exponential tails for bounded averages.}
Chebyshev gives a $1/k^{2}$ tail from the variance alone. When the
summands are bounded and independent, far stronger
\engterm{concentration inequalities} give tails that shrink
exponentially in $n$. \engterm{Hoeffding's inequality} states that for
independent $X_{i} \in [a_{i}, b_{i}]$ with sample mean $\bar X_{n}$,
\[
\Pr\!\left( |\bar X_{n} - \E[\bar X_{n}]| \geq \varepsilon \right)
\leq 2 \exp\!\left( -\frac{2 n^{2} \varepsilon^{2}}
{\sum_{i=1}^{n} (b_{i} - a_{i})^{2}} \right).
\]
For Bernoulli summands on $[0, 1]$ the bound reads $2 e^{-2 n
\varepsilon^{2}}$, which is the engineering tool for sample-size
planning when the only structural fact is boundedness. The proof is
the \engterm{Chernoff method}: apply Markov to $e^{t S_{n}}$, bound
each factor's MGF, and optimise over $t > 0$. The Chernoff bound on a
$\mathrm{Bin}(n, p)$ count, $\Pr(X \geq (1 + \delta) np) \leq
[e^{\delta}/(1 + \delta)^{1 + \delta}]^{np}$, is the workhorse of
randomised-algorithm analysis and of load-balancing arguments, where
it certifies that the busiest server of $m$ servers fed $m$ jobs
carries $O(\log m / \log\log m)$ jobs with high probability. The
practical lesson is the gap between the bounds: to certify that an
empirical defect rate is within $0.01$ of the truth with probability
$0.99$, Chebyshev demands $n$ of order $10^{6}$, while Hoeffding,
using boundedness, demands $n \approx \ln(2/0.01)/(2 \cdot 0.01^{2})
\approx 2.6 \times 10^{4}$, two orders of magnitude fewer samples.
Concentration is what makes finite-sample guarantees, as opposed to
asymptotic CLT statements, available to the engineer who needs a
number before the limit.
\end{mastery}

\subsection{Random walks and the gambler's ruin}

The running sum $S_{n} = \sum_{i=1}^{n} \xi_{i}$ of iid steps is the
simplest non-trivial stochastic process and the one whose intuition an
engineer reaches for whenever a quantity accumulates increments: a
buffer level under random arrivals and departures, a cumulative
measurement error, a financial balance, a particle's displacement. Take
the symmetric simple random walk, $\xi_{i} = \pm 1$ with probability
$\tfrac12$ each, started at $S_{0} = 0$. The walk has mean $\E[S_{n}] =
0$ at every step, yet it does not stay near zero: its standard deviation
grows as $\sqrt{n}$, so the typical displacement after $n$ steps is of
order $\sqrt{n}$. The walk is recurrent in one and two dimensions (it
returns to the origin with probability one) and transient in three or
more, a dimensional dichotomy that governs whether a diffusing
contaminant reliably revisits a sensor and whether a randomised search
reliably revisits a state.

\paragraph{Worked example: gambler's ruin and the absorbing boundary.}
A control buffer holds between $0$ and $N$ tokens; each tick it gains a
token with probability $p$ or loses one with probability $q = 1 - p$,
and the process stops if the buffer empties ($0$) or overflows ($N$).
Starting from $k$ tokens, what is the probability the buffer overflows
before it empties? Let $r_{k}$ be that probability. Conditioning on the
first step gives the recurrence $r_{k} = p\,r_{k+1} + q\,r_{k-1}$ with
boundary values $r_{0} = 0$ and $r_{N} = 1$. For the fair case $p = q =
\tfrac12$ the recurrence makes $r_{k}$ linear, $r_{k} = k/N$: the
overflow probability is just the fractional starting position. For the
biased case the solution is
\[
r_{k} = \frac{1 - (q/p)^{k}}{1 - (q/p)^{N}},
\]
which concentrates sharply once $p \neq q$. With a slight upward drift
$p = 0.52$, $q = 0.48$, and $N = 100$, a buffer starting half full
($k = 50$) overflows before emptying with probability $r_{50} =
[1 - (0.48/0.52)^{50}]/[1 - (0.48/0.52)^{100}] \approx
[1 - 0.0208]/[1 - 0.000432] \approx 0.979$: a four-percentage-point
drift makes overflow nearly certain from the midpoint, where the fair
walk would split fifty-fifty. The expected time to absorption from $k$
in the fair case is $k(N - k)$ steps, which peaks at the midpoint and
sets how long a balanced buffer wanders before it hits a boundary. The
gambler's-ruin arithmetic is the backbone of first-passage analysis: the
sequential-test stopping rule, the time for a random walk to escape a
tolerance band, and the ruin probability of a risk reserve all read off
the same recurrence.

\begin{mastery}
\textbf{Martingales and the optional-stopping theorem.} A sequence
$\{M_{n}\}$ is a \engterm{martingale} with respect to the information
$\mathcal{F}_{n}$ up to step $n$ when $\E[|M_{n}|] < \infty$ and
$\E[M_{n+1} \mid \mathcal{F}_{n}] = M_{n}$: the best forecast of
tomorrow's value, given everything known today, is today's value. A
fair game, the running likelihood ratio under the null hypothesis, and
the conditional-expectation process $\E[Y \mid \mathcal{F}_{n}]$ of a
fixed $Y$ are all martingales. The \engterm{optional-stopping theorem}
states that under mild conditions (a bounded stopping time, or a bounded
martingale) the expected value at a random stopping time $\tau$ equals
the starting value, $\E[M_{\tau}] = \E[M_{0}]$. The theorem solves the
gambler's-ruin times in two lines: $S_{n}$ itself is a martingale in the
fair case, so $\E[S_{\tau}] = 0 = N r_{k} + 0 \cdot (1 - r_{k}) -
\text{(start } k)$ rearranges to $r_{k} = k/N$, and the process $S_{n}^{2}
- n$ is also a martingale, so optional stopping on it yields the
expected absorption time $k(N - k)$ without solving a second recurrence.
Martingales are the rigorous engine behind sequential analysis (Wald's
identity and the sequential probability ratio test), behind the
convergence proofs for stochastic-approximation and reinforcement-
learning algorithms, and behind the concentration inequalities for
dependent data (the Azuma-Hoeffding inequality bounds a martingale with
bounded increments exactly as Hoeffding bounds an independent sum). The
defining property, that conditioning on the past does not change the
expected next value, is the formal statement of ``no predictable trend
to exploit'', which is why martingales model fair games, efficient
markets, and well-calibrated forecasts alike.
\end{mastery}

\section{Stochastic processes preview}\label{sec:vol02:ch13:processes}\pathtag{standard}

A \engterm{stochastic process} is a collection $\{X_{t}\}_{t \in T}$
of random variables on a common probability space, indexed by a set
$T$. The index $T$ is typically a time set: $T = \N$ for
discrete-time processes, $T = [0, \infty)$ for continuous-time. The
joint distribution of any finite subcollection of the $X_{t}$ is
the \engterm{finite-dimensional distribution}; together, these
distributions specify the process.

This section is a preview. Volume IX (systems and control) and
Volume X (failure and reliability) develop the apparatus. We here
treat one canonical example, the discrete-time Markov chain on a
finite state space, and use it to ground the project of
section~\ref{sec:vol02:ch13:project}
\cite[ch.~6]{text:wasserman2004,text:v2ch13-norris-markov}.

\begin{mastery}
\textbf{The Poisson process and its calculus.} The continuous-time
counterpart of the rare-event count is the \engterm{Poisson
process} of rate $\lambda$: a counting process $\{N(t)\}_{t \geq 0}$
with independent increments, $N(0) = 0$, and $N(t) - N(s) \sim
\mathrm{Pois}(\lambda(t - s))$ for $s < t$. The inter-arrival times
are iid $\mathrm{Exp}(\lambda)$, which is the bridge between the
discrete Poisson count and the continuous exponential wait. Two
operations recur in engineering. \emph{Superposition}: merging
independent Poisson processes of rates $\lambda_{1}, \lambda_{2}$
yields a Poisson process of rate $\lambda_{1} + \lambda_{2}$, so the
aggregate arrival stream at a server fed by many independent sources
is itself Poisson. \emph{Thinning}: keeping each arrival
independently with probability $q$ yields a Poisson process of rate
$q\lambda$, so a sampled or filtered arrival stream stays Poisson.
These closure properties are why the Poisson process is the default
model for arrivals in queueing, photon counts in detectors, and
request streams in networked systems, and why the M/M/1 queue of the
design exercises is analytically tractable.
\end{mastery}

\paragraph{Worked example: a Poisson process from arrivals to risk.}
A maintenance depot logs equipment alarms as a Poisson process of
rate $\lambda = 3$ per day. Four questions exercise the standard
toolkit. First, the count in a fixed window: the number of alarms in
a five-day stretch is $\mathrm{Pois}(15)$, with mean $15$ and standard
deviation $\sqrt{15} \approx 3.9$, and the probability of a quiet
five-day window with no alarm is $e^{-15} \approx 3 \times 10^{-7}$.
Second, the wait to the next alarm: inter-arrivals are
$\mathrm{Exp}(3\,\text{day}^{-1})$, so the probability of no alarm for
a full day is $e^{-3} \approx 0.050$, and the median gap is
$\ln 2 / 3 \approx 0.23\,\text{day}$, about five and a half hours.
Third, the wait to the third alarm: the sum of three independent
exponential gaps is $\mathrm{Gamma}(3, 3)$, with mean $3/3 = 1\,
\text{day}$ and variance $3/9 = 0.33\,\text{day}^{2}$, the result the
moment-generating-function derivation above supplies directly. Fourth,
thinning by severity: if each alarm is independently critical with
probability $q = 0.1$, the critical alarms form a Poisson process of
rate $q\lambda = 0.3$ per day, so critical alarms arrive on average
once every $3.3$ days and the count of critical alarms in a $30$-day
month is $\mathrm{Pois}(9)$. The four answers come from one model and
four of its closure properties, which is the reason the Poisson
process is the first reach for any stream of rare independent events
and the analytic backbone of the queueing models in the design
exercises and in Volume IX.

\subsection{Markov chains: definition}

A \engterm{discrete-time Markov chain} on a finite state space
$S = \{1, 2, \ldots, k\}$ is a sequence of random variables
$X_{0}, X_{1}, X_{2}, \ldots$ taking values in $S$, satisfying the
\engterm{Markov property}:
\[
\Pr(X_{n+1} = j \mid X_{n} = i, X_{n-1}, \ldots, X_{0})
= \Pr(X_{n+1} = j \mid X_{n} = i).
\]
Given the present, the future is conditionally independent of the
past. The Markov property is what makes the chain tractable; it
fails for processes with memory beyond one step, and the engineer
who imposes a Markov assumption when the system has long memory
has a subtle modelling error that often shows up only in long-run
behaviour.

The chain is \engterm{time-homogeneous} when the transition
probability $\Pr(X_{n+1} = j \mid X_{n} = i) = P_{ij}$ does not
depend on $n$. We restrict to the time-homogeneous case throughout.
The matrix $\mat{P}$ with entries $P_{ij}$ is the
\engterm{transition matrix}. Each row sums to one; $\mat{P}$ is a
\engterm{row-stochastic matrix}.

\subsection{Multi-step transitions and stationary distributions}

The $n$-step transition probability $\Pr(X_{n} = j \mid X_{0} = i)$
is the $(i, j)$ entry of $\mat{P}^{n}$. The proof is by induction
on $n$ and the law of total probability. A distribution
$\vect{\pi}$ on $S$ (a row vector with $\pi_{i} \geq 0$ and
$\sum_{i} \pi_{i} = 1$) is the \engterm{distribution of $X_{n}$}
when $\Pr(X_{n} = i) = \pi_{i}$. Given $X_{n}$ has distribution
$\vect{\pi}_{n}$, $X_{n+1}$ has distribution $\vect{\pi}_{n+1} =
\vect{\pi}_{n} \mat{P}$.

A distribution $\vect{\pi}$ is \engterm{stationary} for $\mat{P}$
when $\vect{\pi} \mat{P} = \vect{\pi}$. Stationary distributions are
the left eigenvectors of $\mat{P}$ corresponding to the eigenvalue
$1$, normalised to sum to one. Volume II Chapter 10 (eigenvalues
and decompositions) gave the apparatus.

A chain is \engterm{irreducible} when every state can be reached
from every other in some finite number of steps. A chain is
\engterm{aperiodic} when, for some state $i$, the set
$\{n : (\mat{P}^{n})_{ii} > 0\}$ has greatest common divisor $1$
(equivalently, for irreducible chains, every state has this
property). The \engterm{Fundamental Theorem of Markov Chains}
states that a finite, irreducible, aperiodic Markov chain has a
unique stationary distribution $\vect{\pi}$, and for every initial
distribution $\vect{\pi}_{0}$,
\[
\vect{\pi}_{0} \mat{P}^{n} \to \vect{\pi} \quad\text{as }
n \to \infty.
\]
The chain forgets its initial distribution at a rate set by the
second-largest eigenvalue of $\mat{P}$. The \engterm{spectral gap}
$1 - |\lambda_{2}|$ governs the mixing time of the chain
\cite[ch.~6]{text:wasserman2004}.

\subsection{Two examples}

A weather model has two states, sunny ($1$) and rainy ($2$), with
transition matrix
\[
\mat{P} = \begin{pmatrix} 0.9 & 0.1 \\ 0.5 & 0.5 \end{pmatrix}.
\]
The stationary distribution solves $\pi_{1} \cdot 0.9 + \pi_{2}
\cdot 0.5 = \pi_{1}$ with $\pi_{1} + \pi_{2} = 1$, giving
$\pi_{1} = 5/6 \approx 0.833$ and $\pi_{2} = 1/6 \approx 0.167$.
In the long run, about $83\,\%$ of days are sunny.

A simple random walk on the integers $\{0, 1, \ldots, k\}$ moves
$+1$ or $-1$ with probability $1/2$ each, with reflecting
boundaries at $0$ and $k$. The chain is irreducible and aperiodic;
the stationary distribution is uniform. The mixing time grows like
$k^{2}$, which the project of
section~\ref{sec:vol02:ch13:project} asks the reader to verify
empirically.

\paragraph{Worked example: a three-state maintenance chain.}
A machine cycles among three states each shift: running ($1$),
degraded ($2$), and under repair ($3$), with transition matrix
\[
\mat{P} = \begin{pmatrix}
0.90 & 0.08 & 0.02 \\
0.00 & 0.70 & 0.30 \\
0.60 & 0.00 & 0.40
\end{pmatrix}.
\]
Each row sums to one. The stationary distribution solves
$\vect{\pi}\mat{P} = \vect{\pi}$ with $\sum \pi_{i} = 1$. Writing
the three balance equations,
\[
\pi_{1} = 0.90\pi_{1} + 0.60\pi_{3}, \quad
\pi_{2} = 0.08\pi_{1} + 0.70\pi_{2}, \quad
\pi_{3} = 0.02\pi_{1} + 0.30\pi_{2} + 0.40\pi_{3}.
\]
The first gives $0.10\pi_{1} = 0.60\pi_{3}$, so $\pi_{1} =
6\pi_{3}$. The second gives $0.30\pi_{2} = 0.08\pi_{1}$, so
$\pi_{2} = (0.08/0.30)\pi_{1} = 0.2667 \cdot 6\pi_{3} = 1.6\pi_{3}$.
Normalising, $\pi_{1} + \pi_{2} + \pi_{3} = (6 + 1.6 + 1)\pi_{3} =
8.6\pi_{3} = 1$, so $\pi_{3} = 0.1163$, $\pi_{2} = 0.1860$,
$\pi_{1} = 0.6977$. In the long run the machine runs about
$70\,\%$ of shifts, is degraded $19\,\%$, and is under repair
$12\,\%$. The long-run availability is $\pi_{1} + \pi_{2} = 0.884$
if degraded operation still produces output, or $\pi_{1} = 0.698$
if only the running state counts. The same three numbers, multiplied
by a per-shift repair cost and a per-shift lost-production cost,
give the expected steady-state operating cost that a maintenance
contract is priced against. The transition matrix and its stationary
distribution are tabulated in
\repopath{volumes/02-form/ch13-probability/data/maintenance_chain.csv}
for the reader who wants a worked non-trivial chain to check a
solver against. Figure~\ref{fig:vol02:ch13:markov-chain} gives the
weather chain's transition graph and a stylised convergence trace
of the running occupancy fraction toward the stationary value
$\pi_{1} = 5/6$.

\input{volumes/02-form/ch13-probability/figures/fig-markov-chain.tex}

\subsection{LLN for Markov chains}

For an irreducible aperiodic finite-state Markov chain with
stationary distribution $\vect{\pi}$ and any function $g : S \to \R$,
\[
\frac{1}{n} \sum_{t=1}^{n} g(X_{t}) \to \sum_{i \in S}
\pi_{i} g(i) = \E_{\vect{\pi}}[g(X)] \quad\text{a.s.}
\]
The time average of $g$ along a sample path equals the space
average of $g$ under the stationary distribution. The chain is
\engterm{ergodic}, which is the technical name for the property
that a single long sample path explores the state space with the
right relative frequencies. Ergodicity is what licences the project
of section 13.10: a long simulation of the chain produces an
empirical distribution that, by the LLN, converges to the
stationary distribution.

\begin{mastery}
\textbf{Detailed balance and reversibility.} A chain is
\engterm{reversible} with respect to a distribution $\vect{\pi}$
when it satisfies the \engterm{detailed-balance} condition
$\pi_{i} P_{ij} = \pi_{j} P_{ji}$ for every pair $(i, j)$. Summing
over $i$ recovers stationarity, $\sum_{i} \pi_{i} P_{ij} = \pi_{j}
\sum_{i} P_{ji} = \pi_{j}$, so any $\vect{\pi}$ satisfying detailed
balance is automatically stationary. The condition is far easier to
check than the full eigenvector equation because it is local: it
involves only the two states linked by each transition. Reversibility
is the engine behind \engterm{Markov chain Monte Carlo}: the
Metropolis-Hastings algorithm builds a chain whose detailed-balance
distribution is a target the analyst wants to sample, then runs the
chain long enough to draw from it. Volume VII's computational
chapters use exactly this construction. Not every chain is
reversible; the three-state maintenance chain above is not, because
repair returns directly to running but running never returns
directly from repair, so $\pi_{1} P_{13} \neq \pi_{3} P_{31}$ in
general. Reversible chains have real eigenvalues and a clean spectral
theory; non-reversible chains can mix faster but lose that
guarantee.
\end{mastery}

\subsection{Continuous-time chains and birth-death processes}

The discrete-time chain steps on a clock. Many engineering systems
instead change state at random instants: a machine fails at an
exponentially distributed time, a queue gains a customer at a Poisson
instant, a population gains or loses a member. The
\engterm{continuous-time Markov chain} (CTMC) captures these. The
process holds in state $i$ for an $\mathrm{Exp}(q_{i})$ time, then jumps
to state $j \neq i$ with probability $q_{ij}/q_{i}$, where the
\engterm{transition rates} $q_{ij} \geq 0$ ($i \neq j$) are collected
with the diagonal convention $q_{ii} = -q_{i} = -\sum_{j \neq i} q_{ij}$
into the \engterm{generator matrix} $\mat{Q}$, whose rows sum to zero.
The stationary distribution solves the \engterm{global balance}
equations $\vect{\pi}\mat{Q} = \vect{0}$ with $\sum_{i}\pi_{i} = 1$, the
continuous-time analogue of $\vect{\pi}\mat{P} = \vect{\pi}$, and it
reads as a flow conservation law: at stationarity, the probability flux
out of every state equals the flux in.

A \engterm{birth-death process} is the special case where jumps move only
to neighbouring states: from state $n$ the process moves up to $n + 1$
at \engterm{birth rate} $\lambda_{n}$ and down to $n - 1$ at
\engterm{death rate} $\mu_{n}$. The birth-death structure makes the
balance equations solvable in closed form because the flow across the
cut between states $n$ and $n + 1$ must balance in steady state,
$\pi_{n}\lambda_{n} = \pi_{n+1}\mu_{n+1}$ (\engterm{detailed balance},
which every birth-death chain automatically satisfies). Unrolling the
recursion,
\[
\pi_{n} = \pi_{0}\prod_{k=1}^{n} \frac{\lambda_{k-1}}{\mu_{k}},
\qquad
\pi_{0} = \left(1 + \sum_{n=1}^{N}\prod_{k=1}^{n}
\frac{\lambda_{k-1}}{\mu_{k}}\right)^{-1}.
\]
The M/M/1 queue is the birth-death process with constant rates
$\lambda_{n} = \lambda$ and $\mu_{n} = \mu$, giving the geometric
stationary law $\pi_{n} = (1 - \rho)\rho^{n}$ with $\rho = \lambda/\mu$
that the design exercises use. The repair model of the case study below
is the birth-death process read in the language of reliability, with
failures as deaths and repairs as births.

\section{Case study: a repairable redundant system, end to end}
\label{sec:vol02:ch13:availability}\pathtag{standard}

The load-strength case study that follows treats a single mission: the
part either holds or it does not. Many systems instead run, fail, get
repaired, and run again, and the question is not a one-shot reliability
but a long-run \engterm{availability}, the fraction of time the system
is up. This case study works that problem end to end on a two-unit
redundant system with repair, using the continuous-time-chain apparatus
just introduced, then cross-checks the analytical answer with a Monte
Carlo simulation that demonstrates a variance-reduction technique. It is
deliberately a different shape of problem from the load-strength study:
a process unfolding in time, solved by a generator matrix rather than a
tail integral.

\subsection{The model}

A system carries two identical units in active redundancy; the system is
up as long as at least one unit works. Each working unit fails
independently at constant hazard $\lambda$ (so its time to failure is
$\mathrm{Exp}(\lambda)$, the flat middle of the bathtub curve of section
13.4). A single repair crew restores a failed unit at rate $\mu$ (repair
time $\mathrm{Exp}(\mu)$); if both units are down, they are repaired one
at a time. The state is the number of \emph{failed} units, $n \in
\{0, 1, 2\}$, and the process is a birth-death chain with failures as
births and repairs as deaths. From state $0$ (both up) either unit can
fail, so the upward rate is $2\lambda$. From state $1$ (one up, one
under repair) the single working unit fails at rate $\lambda$ while the
crew completes a repair at rate $\mu$. From state $2$ (both down) the
crew repairs at rate $\mu$. The generator is
\[
\mat{Q} = \begin{pmatrix}
-2\lambda & 2\lambda & 0 \\
\mu & -(\lambda + \mu) & \lambda \\
0 & \mu & -\mu
\end{pmatrix},
\]
each row summing to zero. State $2$ is the system-down state; states $0$
and $1$ are up.

\subsection{Solving for availability}

Detailed balance across the two cuts gives the stationary distribution
directly. The cut between $0$ and $1$ balances $\pi_{0}\cdot 2\lambda =
\pi_{1}\cdot\mu$, so $\pi_{1} = (2\lambda/\mu)\pi_{0}$. The cut between
$1$ and $2$ balances $\pi_{1}\cdot\lambda = \pi_{2}\cdot\mu$, so
$\pi_{2} = (\lambda/\mu)\pi_{1} = (2\lambda^{2}/\mu^{2})\pi_{0}$.
Writing $r = \lambda/\mu$ for the failure-to-repair rate ratio and
normalising,
\[
\pi_{0} = \frac{1}{1 + 2r + 2r^{2}}, \quad
\pi_{1} = \frac{2r}{1 + 2r + 2r^{2}}, \quad
\pi_{2} = \frac{2r^{2}}{1 + 2r + 2r^{2}}.
\]
The \engterm{steady-state availability} is the fraction of time the
system is up, $A = \pi_{0} + \pi_{1} = 1 - \pi_{2}$, and the
\engterm{unavailability} is
\[
\bar A = \pi_{2} = \frac{2r^{2}}{1 + 2r + 2r^{2}}.
\]
The quadratic in $r$ is the signature of the redundancy: unavailability
scales as $r^{2}$ for small $r$, so two units cut the downtime to the
square of a single unit's downtime, which is the entire quantitative
case for active redundancy with repair.

\subsection{Worked numbers}

A unit has mean time to failure $1/\lambda = 2000\,\text{h}$ and mean
time to repair $1/\mu = 8\,\text{h}$, so $r = \lambda/\mu = 8/2000 =
0.004$. The unavailability is
\[
\bar A = \frac{2(0.004)^{2}}{1 + 2(0.004) + 2(0.004)^{2}}
= \frac{3.2 \times 10^{-5}}{1.008032}
\approx 3.17 \times 10^{-5},
\]
an availability of $A \approx 0.9999683$, about $0.28\,\text{h}$ of
downtime per year. A single unit would be down a fraction $r/(1 + r) =
0.004/1.004 \approx 3.98 \times 10^{-3}$ of the time, roughly
$34.9\,\text{h}$ per year; the second unit and the repair crew buy two
orders of magnitude. The dependence on the repair rate is steep: if the
crew is slowed so the mean repair time triples to $24\,\text{h}$, then
$r = 0.012$ and $\bar A \approx 2(0.012)^{2}/1.0243 \approx 2.81 \times
10^{-4}$, a ninefold rise in downtime from a threefold rise in repair
time, the squared sensitivity again. Availability engineering buys more
from fast repair than from reliable units, because the redundancy turns
the repair-time ratio into the dominant lever.

\subsection{The single-crew bottleneck and the two-crew variant}

The model above shares one repair crew, which is why the rate out of
state $2$ is $\mu$ and not $2\mu$. With a second crew, both failed units
are repaired in parallel, the downward rate from state $2$ becomes
$2\mu$, and the state-$2$ balance becomes $\pi_{1}\lambda = \pi_{2}\cdot
2\mu$, so $\pi_{2} = (r/2)\pi_{1} = r^{2}\pi_{0}$. The unavailability
falls to $\bar A_{2} = r^{2}/(1 + 2r + r^{2}) = r^{2}/(1 + r)^{2}$,
which for the base numbers is $(0.004)^{2}/(1.004)^{2} \approx 1.59
\times 10^{-5}$, half the single-crew figure. The factor of two is the
value of the second crew, and it is small precisely because state $2$ is
already rare: the crew that matters is the one that handles the first
failure quickly, not the one that handles the unlikely double failure.
An availability budget that spends on a second crew before it spends on
faster first-failure response has misread where the downtime lives, a
conclusion the model delivers and a deterministic redundancy count
cannot.

\subsection{Monte Carlo cross-check with a control variate}

The closed form is available here because the chain is small and a
birth-death process. For a larger system (more units, non-exponential
repair, a shared spare-parts pool) the generator may have no tidy
solution, and the engineer simulates. We cross-check the analytical
availability by simulating the CTMC: from each state, draw the holding
time as the minimum of the competing exponential clocks, jump to the
winning state, and accumulate time-in-state until a long horizon $T$.
The estimated availability is the time-weighted fraction of the run
spent in states $0$ and $1$. Because availability is near one, the
\emph{unavailability} is the rare quantity, and a raw estimator of
$\bar A \approx 3 \times 10^{-5}$ from a finite run has a large relative
error: the relative standard error of a fraction-$p$ estimator is
$\sqrt{(1 - p)/(pT\mu)}$-like, blowing up as $p \to 0$, the rare-event
difficulty the failure section of Volume I Chapter 8 named.

A \engterm{control variate} cuts that variance. The trick is to subtract
a correlated quantity whose mean is known and add the mean back. Here the
time spent in state $1$ (single failure) is strongly correlated with the
time spent in state $2$ (double failure), because both are driven by the
same failure stream, and the state-$1$ occupancy has a known analytical
mean $\pi_{1}$. The control-variate estimator of unavailability is
\[
\hat{\bar A}_{\mathrm{cv}} = \hat\pi_{2}
- c\,(\hat\pi_{1} - \pi_{1}),
\]
where $\hat\pi_{1}, \hat\pi_{2}$ are the simulated occupancies, $\pi_{1}$
is the exact value, and the coefficient $c = \Cov(\hat\pi_{2},
\hat\pi_{1})/\Var(\hat\pi_{1})$ is chosen to minimise variance (estimated
from the runs themselves). The estimator is unbiased because the
subtracted term has expectation zero, and its variance is
$\Var(\hat\pi_{2})(1 - \rho^{2}_{12})$, where $\rho_{12}$ is the
correlation between the two occupancy estimates: the closer the control
variate tracks the target, the larger the variance cut. For this chain
the two occupancies correlate strongly, and the control variate reduces
the number of simulated transitions needed for a given relative error by
roughly an order of magnitude, the same kind of payoff the
square-root-of-$n$ wall principle promised from a better estimator
rather than more samples. The companion script
\repopath{volumes/02-form/ch13-probability/code/availability_ctmc.py}
builds the generator, solves $\vect{\pi}\mat{Q} = \vect{0}$, simulates
the CTMC, and reports the raw and control-variate unavailability
estimates against the closed form, for the single-crew and two-crew
variants and for parameters of the reader's choice. The transition
rates and the analytical stationary distribution are tabulated in
\repopath{volumes/02-form/ch13-probability/data/availability_chain.csv}.
Figure~\ref{fig:vol02:ch13:availability} draws the three-state
birth-death graph and the unavailability $\bar A(r)$ as a function of
the failure-to-repair ratio, with the single-crew and two-crew curves
and their $r^{2}$ small-ratio asymptotes.

\input{volumes/02-form/ch13-probability/figures/fig-availability.tex}

\subsection{Reading the result}

The availability study delivers habits that the one-shot reliability of
the load-strength study cannot. Long-run behaviour of a repairable
system is a stationary distribution of a continuous-time chain, read off
a generator matrix, not a single failure integral. Redundancy with
repair turns the failure-to-repair ratio $r$ into the lever, and
unavailability scales as $r^{2}$, so fast repair and fast first-failure
response dominate the budget. The closed form is a gift of the
birth-death structure; when it is absent, simulation answers the same
question, and a control variate built from a correlated quantity with
known mean buys back the precision that rare-event estimation loses. The
two case studies together cover the two faces of probabilistic
engineering: the tail of a static margin, and the steady state of a
dynamic process.

\section{Case study: load-strength interference}\label{sec:vol02:ch13:case-study}\pathtag{standard}

The apparatus of this chapter, distributions, transformations,
expectation, the normal tail, and the propagation of uncertainty,
converges on one of the oldest quantitative questions in engineering:
will the part hold the load? The probabilistic answer is the
\engterm{load-strength interference} model, also called the
stress-strength model, and working it end to end ties the chapter
together on a problem an engineer actually ships against
\cite[ch.~5]{text:ross-probability}.

\subsection{The model}

A component has a random strength $S$ and is subjected to a random
load $L$, both in the same units (a stress, a force, a pressure). The
component survives the mission when $S > L$ and fails when $L \geq S$.
The \engterm{probability of failure} is
\[
p_{f} = \Pr(L \geq S) = \Pr(S - L \leq 0),
\]
and the \engterm{reliability} is $R = 1 - p_{f}$. The deterministic
design tradition replaces the two distributions by single numbers, a
rated strength and a rated load, and inserts a \engterm{safety factor}
$\mathrm{SF} = \bar S / \bar L$ to cover the gap. The probabilistic
model exposes what the safety factor hides: two designs with the same
nominal safety factor can have failure probabilities that differ by
orders of magnitude, because the safety factor says nothing about the
spread of either distribution. The interference is the overlap of the
two densities, drawn in
Figure~\ref{fig:vol02:ch13:load-strength}, and the failure probability
is the probability mass in that overlap region, not the distance
between the means.

\input{volumes/02-form/ch13-probability/figures/fig-load-strength.tex}

\subsection{The Gaussian case, in full}

Take both load and strength Gaussian and independent:
$L \sim \mathcal{N}(\mu_{L}, \sigma_{L}^{2})$ and $S \sim
\mathcal{N}(\mu_{S}, \sigma_{S}^{2})$. The \engterm{margin}
$M = S - L$ is a linear combination of independent Gaussians, so by
the closure property it is itself Gaussian,
\[
M \sim \mathcal{N}\!\left( \mu_{S} - \mu_{L},\
\sigma_{S}^{2} + \sigma_{L}^{2} \right).
\]
Failure is the event $M \leq 0$. Standardising,
\[
p_{f} = \Pr(M \leq 0) = \Phi\!\left(
\frac{0 - (\mu_{S} - \mu_{L})}
{\sqrt{\sigma_{S}^{2} + \sigma_{L}^{2}}} \right)
= \Phi(-\beta), \qquad
\beta = \frac{\mu_{S} - \mu_{L}}
{\sqrt{\sigma_{S}^{2} + \sigma_{L}^{2}}}.
\]
The dimensionless quantity $\beta$ is the \engterm{reliability index}:
the number of standard deviations of the margin by which the mean
strength exceeds the mean load. It is the single number that sets the
failure probability in the Gaussian model, and it is the quantity the
load-and-resistance-factor design codes are calibrated against. The
mapping from $\beta$ to $p_{f}$ is the standard-normal tail: $\beta =
3$ gives $p_{f} = \Phi(-3) \approx 1.35 \times 10^{-3}$, $\beta = 4$
gives $\approx 3.17 \times 10^{-5}$, and $\beta = 5$ gives $\approx
2.87 \times 10^{-7}$. Each extra unit of $\beta$ buys roughly a factor
of thirty in reliability across this range.

\subsection{Worked numbers}

A pressure vessel wall has burst strength with mean $\mu_{S} = 600\,
\text{bar}$ and standard deviation $\sigma_{S} = 40\,\text{bar}$,
reflecting material and weld-quality scatter. The operating load has
mean $\mu_{L} = 400\,\text{bar}$ and standard deviation $\sigma_{L} =
30\,\text{bar}$, reflecting pressure-cycle variation. The nominal
safety factor is $\mathrm{SF} = 600/400 = 1.5$. The reliability index
is
\[
\beta = \frac{600 - 400}{\sqrt{40^{2} + 30^{2}}}
= \frac{200}{\sqrt{1600 + 900}}
= \frac{200}{50} = 4.0,
\]
so the failure probability is $p_{f} = \Phi(-4.0) \approx 3.2 \times
10^{-5}$, about three failures per hundred thousand vessels. Now hold
the safety factor fixed at $1.5$ but suppose the weld process degrades
and $\sigma_{S}$ rises to $80\,\text{bar}$. The means are unchanged,
so the deterministic safety factor still reads $1.5$ and a
factor-of-safety acceptance test passes. The reliability index falls
to
\[
\beta = \frac{200}{\sqrt{80^{2} + 30^{2}}}
= \frac{200}{\sqrt{6400 + 900}}
= \frac{200}{85.4} = 2.34,
\]
and the failure probability climbs to $\Phi(-2.34) \approx 9.6 \times
10^{-3}$, a factor of three hundred worse than the original design at
the same safety factor. The lesson the deterministic number cannot
carry is that scatter in strength is as load-bearing as the mean: a
process-control lapse that doubles the strength standard deviation
costs three hundredfold in reliability while leaving every nominal
margin intact.

\subsection{The lognormal case and why it is often the right one}

Strength and load are non-negative and frequently right-skewed, which
makes the lognormal a better model than the normal for many
materials. When $S$ and $L$ are lognormal, $\ln S$ and $\ln L$ are
normal, and the failure event $S > L$ is the same as $\ln S > \ln L$,
so the margin in log space, $\ln S - \ln L$, is normal. The reliability
index becomes
\[
\beta = \frac{\mu_{\ln S} - \mu_{\ln L}}
{\sqrt{\sigma_{\ln S}^{2} + \sigma_{\ln L}^{2}}},
\]
with the log-space parameters obtained from the means and coefficients
of variation by $\sigma_{\ln}^{2} = \ln(1 + \mathrm{CV}^{2})$ and
$\mu_{\ln} = \ln \mu - \tfrac12 \sigma_{\ln}^{2}$. The lognormal model
changes the answer most in the tail, which is exactly where the failure
probability lives. For the original vessel numbers, treating the
coefficients of variation $\mathrm{CV}_{S} = 40/600 = 0.067$ and
$\mathrm{CV}_{L} = 30/400 = 0.075$ as lognormal gives $\sigma_{\ln S}
\approx 0.0666$, $\sigma_{\ln L} \approx 0.0748$, and $\beta =
(\ln 600 - \tfrac12 \cdot 0.0666^{2}) - (\ln 400 - \tfrac12 \cdot
0.0748^{2})$ over $\sqrt{0.0666^{2} + 0.0748^{2}} = 0.4035/0.1001
\approx 4.03$, close to the Gaussian $\beta = 4.0$ at these small
coefficients of variation but diverging from it as the scatter grows.
For coefficients of variation above about $0.2$ the two models give
materially different failure probabilities, and the choice between
them is an engineering judgement about the physical skewness of the
strength data, made by the family-selection checklist of section 13.4,
not a matter of convenience.

\subsection{When the inputs are correlated}

The independence of $S$ and $L$ is an assumption, and the failure
section that follows names what happens when it is false. If a common
factor, a temperature excursion that both weakens the material and
raises the internal pressure, couples strength and load with
correlation $\rho$, the margin variance picks up a cross term,
\[
\Var(M) = \sigma_{S}^{2} + \sigma_{L}^{2} - 2\rho\,\sigma_{S}\sigma_{L},
\]
the minus sign appearing because $M = S - L$. Positive correlation
between strength and load (high load coincides with high strength)
shrinks the margin variance and raises $\beta$, improving reliability;
negative correlation (high load coincides with low strength, the
adverse common cause) inflates the margin variance and cuts $\beta$.
For the original vessel at $\rho = -0.5$, $\Var(M) = 1600 + 900 -
2(-0.5)(40)(30) = 2500 + 1200 = 3700$, so $\beta = 200/60.8 = 3.29$
and $p_{f} = \Phi(-3.29) \approx 5 \times 10^{-4}$, fifteen times the
independent figure. An interference analysis that assumes independent
load and strength when a shared driver couples them adversely reports
a reliability the structure does not have, which is the chapter's
governing failure mode read in the language of the case study.

\subsection{Reading the result}

The case study delivers four transferable habits. The failure
probability lives in the overlap of two distributions, not in the gap
between two means, so the spread of each input is a first-class design
variable. A single dimensionless index $\beta$ summarises a Gaussian
interference problem and maps to $p_{f}$ through the normal tail. The
choice of distribution family changes the answer in the tail, which is
the part that matters, so the family-selection discipline of section
13.4 is not academic. And the independence of load and strength is a
claim to be justified, not a default, because an adverse common cause
moves the failure probability by more than a typical design margin. The
companion script
\repopath{volumes/02-form/ch13-probability/code/load_strength.py}
computes $\beta$ and $p_{f}$ for the Gaussian, lognormal, and
correlated cases and runs a Monte Carlo cross-check against the
closed-form tail, so the reader can confirm the interference integral
on an arbitrary pair of input distributions where no closed form
exists.

\section{Failure: independence assumed where there was none}\pathtag{core}

Volume I Chapter 8 §8.7 catalogued the failure modes of statistical
inference under the heading replication crisis. We here take a
parallel pass at the failure modes of probability \emph{modelling}.
The dominant pattern, in finance, in safety analysis, and in
machine learning, is the same: independence assumed across events
that share a common driver, with the consequence that joint failure
probabilities are systematically understated.

\subsection{Tail risk and the joint failure of independent assumptions}

The 2008 financial crisis is the largest documented case of this
failure pattern in the modern engineering literature. The
mortgage-backed-securities pricing models that were standard
through the mid-2000s, including those built on the Gaussian-copula
default-correlation framework of \textcite{paper:v2ch13-li-gaussian-copula},
assumed correlations between mortgage defaults that were calibrated
to a recent low-default era. Those models priced the joint
probability of a large fraction of mortgages defaulting
simultaneously by treating the defaults as nearly independent given
a small number of common factors. When the housing market turned,
the correlations rose sharply: defaults that the models had treated
as conditionally near-independent turned out to share unmodelled
drivers, and the joint failure probabilities the models had assigned
were too low by orders of magnitude. Triple-A-rated tranches built
on the independence assumption defaulted in clusters the models
had assigned probabilities of less than one in ten thousand. The post-crisis literature
\cite{paper:v2ch13-acharya-richardson-2009,text:v2ch13-mackenzie-engine}
identifies mismeasured correlation as a primary driver of the
cluster defaults, and the sociological mechanism by which a
formula known internally to have correlation-sensitivity above some
narrow regime was permitted to underwrite multi-trillion-dollar
exposures.

Figure~\ref{fig:vol02:ch13:fat-tail} renders the deeper companion
error that runs alongside mismeasured correlation: fitting a thin
Gaussian centre and extrapolating it into a tail that is in fact
heavy.

\input{volumes/02-form/ch13-probability/figures/fig-fat-tail.tex}

The pattern is general. When a probability model assigns a small
joint failure probability by appealing to independence, and the
dependence is in fact driven by a hidden common factor, the model's
tail probabilities are too small. The shortfall scales with the
number of failures the model is summing: for $n$ events with
individual probability $p$ and assumed independence, the joint
failure probability is $p^{n}$; if the events share a common
factor, it can be as high as $p$. The ratio is $p^{n - 1}$, a
factor that grows without bound as $n$ grows.

The same pattern appears in:

\begin{itemize}[leftmargin=*]
  \item Reliability analysis of redundant systems, where two
    nominally independent backup channels share a common power
    supply, a common cooling system, or a common firmware version,
    and the assumed independent failure probability is far below
    the actual joint failure probability under shared-cause
    conditions.
  \item Machine-learning ensemble methods that assume independent
    errors across base learners trained on the same data; the
    common training set induces correlated errors that the
    ensemble's variance-reduction argument does not see.
  \item Sensor-fusion architectures that combine multiple sensors
    by treating their noise as independent; environmental
    conditions (temperature, vibration, EMI) that affect every
    sensor produce correlated noise the fusion model misses.
  \item Multiple-hypothesis testing where the test statistics are
    treated as independent for the Bonferroni correction; correlated
    statistics inflate the family-wise error rate beyond what the
    correction controls.
\end{itemize}

A second short illustration: the Northeast blackout of August 2003,
the largest power outage in North American history, propagated
across multiple utilities and grid operators that had each modelled
their own subsystem's failure probability without accounting for the
common factor of grid-wide reactive-power deficiency. Each operator
held a model in which their loads were near-independent of their
neighbours' loads on the timescale relevant to operator action; the
joint event was a $50$-million-customer outage. The post-event
investigation by the U.S.-Canada Task Force documents the
correlation channel (the shared transmission grid) that no single
operator's model represented. The structural lesson is the same as
in 2008: independence treated as a default rather than as a claim
to be justified inflates the apparent safety margin of every
subsystem at once.

\subsection{Diagnostic discipline}

Three habits reduce the risk of this failure mode.

First, name the dependence channel before invoking independence.
What common factor could couple the events the model is treating as
independent? Power supply, environment, training data, calibration
batch, regulatory regime, market regime. If a plausible common
factor exists and the model has no representation of it, the
independence assumption is doing work the data have not earned.

Second, run a sensitivity analysis on the correlation parameter.
What does the model's tail probability look like at correlation
$0.0$, $0.3$, $0.7$, $0.95$? If the answer is qualitatively
different (a factor of ten or more between the extremes), the
model is correlation-sensitive and the engineer who reports a
single number is misleading the recipient.

\paragraph{Worked example: how fast the joint risk moves with
correlation.} Take ten nominally independent channels, each with
failure probability $p = 0.05$, and ask for the probability that at
least eight fail together, the kind of clustered event a redundancy
budget is meant to make negligible. Model the dependence with a
single common factor: each channel fails if a shared latent
Gaussian $Z$ plus an idiosyncratic Gaussian crosses a threshold,
with pairwise correlation $\rho$ between the latent triggers. Under
independence ($\rho = 0$) the probability of eight or more failures
is $\sum_{k=8}^{10} \binom{10}{k} 0.05^{k} 0.95^{10-k} \approx
4 \times 10^{-9}$, comfortably below any sane threshold. At
$\rho = 0.3$ the same eight-of-ten probability rises to roughly
$10^{-5}$; at $\rho = 0.7$ to roughly $10^{-3}$; at $\rho = 0.95$
to a few percent. The span from $\rho = 0$ to $\rho = 0.95$ is six
orders of magnitude on the quantity the safety case turns on. A
report that states $4 \times 10^{-9}$ without naming the
independence assumption has hidden the only number that matters.
The companion script
\repopath{volumes/02-form/ch13-probability/code/correlated_bernoulli.py}
reproduces this sweep for parameters of the reader's choice.

Third, prefer the conservative bound when independence is in doubt.
Section 13.1 gave the union bound, $\Pr(\bigcup A_{i}) \leq \sum
\Pr(A_{i})$, as the bound that does not require independence. For
small individual probabilities the union bound is close to the
sum-of-marginals; the difference between using it and assuming
independence is small for routine problems, and large for
joint-tail problems. An engineer who uses the union bound for the
upper tail and the independence assumption only after explicit
justification has built in a layer of correlation conservatism that
costs little when the assumption holds and saves the analysis when
it does not.

\begin{principle}[Independence is a strong assumption]
Independence is a structural property of the joint distribution.
It does not follow from the marginals being known, from the
covariance being small, or from the sample correlation being
near zero. Independence is licensed by a generative model in which
the events do not share a common cause; in its absence, joint
failure probabilities computed from product-of-marginals are
upper-bounded by the union bound and lower-bounded by the
worst-case dependence, and the truth depends on the dependence
structure the engineer has not measured.
\end{principle}

\section{Project}\label{sec:vol02:ch13:project}\pathtag{core}

\begin{project}[Markov chain simulation and convergence]
Per the project track resolved for this chapter (Q55-vol02-ch13 in
\repopath{docs/editorial-decisions.md}), the artefact is a
Markov-chain simulation of the reader's choice for at least
$10\,000$ steps, with empirical state-occupancy compared against
the analytical stationary distribution and a 1500-word reflection.

\textbf{Scope.}

\begin{itemize}[leftmargin=*]
  \item Choose a Markov chain on a finite state space $S$ with
    $|S|$ between $2$ and $20$ states. Suggested choices: a
    two-state weather model with transition matrix of the reader's
    choice; a five-state random walk on a graph of the reader's
    design; a simple queueing model with a finite buffer; a
    board-game position chain (Snakes and Ladders, simplified
    Monopoly). Document the state space and the transition matrix
    $\mat{P}$ explicitly.
  \item Verify by hand that $\mat{P}$ is row-stochastic
    ($\sum_{j} P_{ij} = 1$ for every $i$) and that the chain is
    irreducible and aperiodic on the chosen state space. If the
    chain is reducible, restrict to a single recurrent class. If
    the chain is periodic, modify $\mat{P}$ by a small lazy
    correction (replace $\mat{P}$ with $(\mat{P} + \mat{I})/2$).
  \item Compute the analytical stationary distribution
    $\vect{\pi}$ as the left eigenvector of $\mat{P}$ corresponding
    to eigenvalue $1$, normalised to sum to one. Verify
    $\vect{\pi} \mat{P} = \vect{\pi}$ by direct multiplication.
\end{itemize}

\textbf{Simulation.}

\begin{itemize}[leftmargin=*]
  \item Implement the chain in a programming environment of the
    reader's choice (Python+NumPy, MATLAB, Julia, R). Seed the
    pseudo-random-number generator and document the seed for
    reproducibility.
  \item Run the chain for at least $N = 10^{4}$ steps from a
    chosen initial state. Record the full trajectory.
  \item Compute the empirical state-occupancy distribution
    $\hat{\vect{\pi}}$ with $\hat\pi_{i} = (1/N) |\{t :
    X_{t} = i\}|$.
  \item Compute the total variation distance between
    $\hat{\vect{\pi}}$ and $\vect{\pi}$,
    $\mathrm{TV}(\hat{\vect{\pi}}, \vect{\pi}) =
    (1/2) \sum_{i} |\hat\pi_{i} - \pi_{i}|$. Report the value.
  \item Plot the running estimate $\hat\pi_{i}^{(t)} = (1/t)
    |\{s \leq t : X_{s} = i\}|$ as a function of $t$ for each
    state $i$, with the analytical $\pi_{i}$ as a horizontal
    reference line. The plot should show convergence at rate
    approximately $1/\sqrt{t}$ around the stationary value.
\end{itemize}

\textbf{Convergence rate and limit theorems.}

\begin{itemize}[leftmargin=*]
  \item Compute the eigenvalues of $\mat{P}$. Identify the second-
    largest eigenvalue magnitude $|\lambda_{2}|$. The mixing time
    is bounded above by a constant times $1/(1 - |\lambda_{2}|)$.
    Report the value and compare against the empirical mixing time
    (the smallest $T$ such that the running estimate from any
    initial state is within $0.05$ in total variation of
    $\vect{\pi}$).
  \item Choose a function $g : S \to \R$ of the reader's choice
    (the indicator of a state, the value of the state under a
    chosen integer labelling, or any function with a non-trivial
    expectation under $\vect{\pi}$). Compute $\E_{\vect{\pi}}[g] =
    \sum_{i} \pi_{i} g(i)$ analytically.
  \item Verify the LLN: plot the running average $(1/t)
    \sum_{s=1}^{t} g(X_{s})$ as a function of $t$, with
    $\E_{\vect{\pi}}[g]$ as a horizontal reference. The running
    average should converge to the reference at rate
    approximately $1/\sqrt{t}$.
  \item Verify the CLT: run $M = 200$ independent copies of the
    chain, each for the same length $N$. Compute the standardised
    sample mean $Z_{m} = \sqrt{N}(\bar g_{m} -
    \E_{\vect{\pi}}[g])/\sigma_{g}$ for each copy, where
    $\sigma_{g}$ is an asymptotic variance that the reader may
    estimate empirically. Plot a histogram of $\{Z_{m}\}$ and
    compare to the standard normal density.
\end{itemize}

\textbf{Deliverable.} A 5-page report comprising:
specification of the chain (state space, transition matrix,
generative interpretation); the analytical stationary distribution
and its derivation; the empirical results (occupancy distribution,
total variation distance, running plots, mixing-time comparison,
LLN and CLT verification); plus a 1500-word reflection on what the
limit theorems guarantee and what they do not, and on when a
Markov assumption is and is not safe in a real engineering system.
The reflection should name at least one real engineering process
the reader has met (in Volume I project work, in their own
practice, or in the reading) and assess honestly whether the Markov
assumption fits.

\smallskip
\textbf{Phases (suggested order, four sessions of about two hours).}

\begin{enumerate}[leftmargin=*]
  \item \emph{Specification.} Choose the chain, write $\mat{P}$,
    verify row-stochasticity, sketch the transition graph by hand,
    confirm irreducibility and aperiodicity.
  \item \emph{Analytical solution.} Solve $\vect{\pi}\mat{P} =
    \vect{\pi}$ algebraically. Verify by direct multiplication. Find
    the eigenvalues of $\mat{P}$ and identify the second-largest
    magnitude.
  \item \emph{Simulation.} Implement the chain. Run for at least
    $10^{4}$ steps from a chosen initial state. Tabulate occupancy.
    The script
    \repopath{volumes/02-form/ch13-probability/code/markov_chain_simulation.py}
    is a working example for the two-state weather chain; the reader
    adapts it to the chosen chain.
  \item \emph{Verification.} Plot the running occupancy and the
    running average of $g$. Run $M = 200$ independent copies for the
    CLT histogram. Tabulate the empirical mixing time and compare it
    to the spectral-gap bound.
  \item \emph{Reflection.} Draft and revise the 1500-word reflection
    on what the limit theorems guarantee, what they do not, and on
    the engineering process from which the chain was abstracted.
\end{enumerate}

\textbf{Success criteria.} The project is complete when each of the
following holds:

\begin{itemize}[leftmargin=*]
  \item The transition matrix is row-stochastic to machine precision
    and is documented with its generative interpretation.
  \item The analytical $\vect{\pi}$ satisfies $\vect{\pi}\mat{P} =
    \vect{\pi}$ to machine precision and sums to one.
  \item The empirical $\hat{\vect{\pi}}$ satisfies
    $\mathrm{TV}(\hat{\vect{\pi}}, \vect{\pi}) < 0.02$ for the chosen
    $N \geq 10^{4}$ (or a documented justification is given for a
    looser bound on a slowly mixing chain).
  \item The running-occupancy plot visibly converges to the
    horizontal reference at a rate consistent with $1/\sqrt{t}$
    fluctuations.
  \item The CLT histogram of $\{Z_{m}\}$ overlays the standard
    normal to the eye, with a one-paragraph note on residual
    discrepancies and their plausible causes.
  \item The reflection identifies a real engineering process, names
    the candidate state space and transition probabilities, and
    states honestly which of irreducibility, aperiodicity,
    time-homogeneity, and the Markov property the process satisfies
    or violates.
\end{itemize}

\textbf{Common failure modes (and their fixes).}

\begin{itemize}[leftmargin=*]
  \item \emph{Periodic chain mistaken for aperiodic.} A simple
    random walk on $\{0, 1, \ldots, k\}$ \emph{without} reflecting
    boundaries is periodic (period $2$); occupancy alternates.
    Fix: add a lazy step ($\mat{P} \mapsto (\mat{P} + \mat{I})/2$)
    or add self-loops to break the period.
  \item \emph{Reducible chain.} A transition matrix with a closed
    subset that the chain enters and never leaves does not have a
    unique stationary distribution; the empirical occupancy depends
    on the initial state. Fix: restrict the analysis to a single
    recurrent class.
  \item \emph{Burn-in not discarded.} Including the first few
    hundred steps in the occupancy estimate skews $\hat{\vect{\pi}}$
    toward the initial distribution. Fix: discard the first
    $5/(1 - |\lambda_{2}|)$ steps as burn-in or run long enough that
    the burn-in is negligible.
  \item \emph{Numerical issues in $\vect{\pi}$.} Solving
    $\vect{\pi}\mat{P} = \vect{\pi}$ as a linear system with a
    single normalisation constraint is numerically stable; solving
    it via direct power iteration on $\mat{P}^{n}$ for large $n$
    can lose precision. Fix: use a standard eigenvector solver and
    renormalise.
  \item \emph{CLT verification at too small an $M$.} Twenty
    independent copies are too few to see a Gaussian shape; the
    histogram looks ragged. Fix: use $M \geq 100$, or report the
    empirical mean and standard deviation of the $\{Z_{m}\}$ and
    compare to $0$ and $1$.
\end{itemize}
\end{project}

\section{Exercises}

The exercises run across the Q16 categories: calculation,
derivation, estimation, simulation, design, and judgement. The
target count is forty. Difficulty rises within each category; the
later items in each subsection are harder.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A fair die is rolled twice. Compute the probability that the sum
of the two rolls is exactly $7$. Compute the probability that the
sum is at least $9$.
\end{exercise}

\paragraph{Solution.}
The sample space is the $36$ equally likely ordered pairs $(i, j)$
with $i, j \in \{1, \ldots, 6\}$. For the sum equal to $7$ the
favourable outcomes are $(1,6), (2,5), (3,4), (4,3), (5,2), (6,1)$,
six in all, so $\Pr(\text{sum} = 7) = 6/36 = 1/6 \approx 0.167$.
Seven is the most likely sum precisely because it has the most
realisations. For the sum at least $9$, count by value: sum $9$ has
$(3,6),(4,5),(5,4),(6,3)$, four outcomes; sum $10$ has
$(4,6),(5,5),(6,4)$, three; sum $11$ has $(5,6),(6,5)$, two; sum
$12$ has $(6,6)$, one. The total is $4 + 3 + 2 + 1 = 10$, so
$\Pr(\text{sum} \geq 9) = 10/36 = 5/18 \approx 0.278$.

\begin{exercise}
A bag contains five red, three blue, and two green balls. Three
balls are drawn without replacement. Compute the probability that
all three are different colours.
\end{exercise}

\paragraph{Solution sketch.}
Number of ways to pick one of each colour: $5 \cdot 3 \cdot 2 = 30$
(choices of which ball of each colour). Number of ordered triples
of three balls is $10 \cdot 9 \cdot 8 = 720$, but the
``one-of-each'' event corresponds to $3! = 6$ orderings of each
unordered triple, so unordered count is $30$ out of $\binom{10}{3}
= 120$. Probability $30/120 = 1/4 = 0.25$. Equivalently, pick three
positions and assign colours: $\binom{5}{1}\binom{3}{1}\binom{2}{1}
/ \binom{10}{3} = 30/120$.

\begin{exercise}
$X \sim \mathrm{Bin}(10, 0.3)$. Compute $\Pr(X = 4)$, $\E[X]$, and
$\Var(X)$.
\end{exercise}

\paragraph{Solution.}
The binomial coefficient is $\binom{10}{4} = 210$. The success
factor is $(0.3)^{4} = 0.0081$ and the failure factor is
$(0.7)^{6} = 0.117649$, so
\[
\Pr(X = 4) = 210 \cdot 0.0081 \cdot 0.117649
= 210 \cdot 0.000953 \approx 0.2001.
\]
The mean is $\E[X] = np = 10 \cdot 0.3 = 3.0$, so the modal count
sits just below the mean, as expected for a right-of-centre $k = 4$.
The variance is $\Var(X) = np(1 - p) = 10 \cdot 0.3 \cdot 0.7 =
2.1$, giving standard deviation $\sigma = \sqrt{2.1} \approx 1.449$.
A useful cross-check: $k = 4$ is $(4 - 3.0)/1.449 \approx 0.69$
standard deviations above the mean, consistent with a probability
near the peak of the distribution.

\begin{exercise}
$X \sim \mathrm{Pois}(2.5)$. Compute $\Pr(X \geq 3)$. State whether
the Markov inequality applied to $X$ at $a = 5$ gives a tighter or
looser bound than the exact value.
\end{exercise}

\paragraph{Solution sketch.}
$\Pr(X = k) = e^{-2.5}\, 2.5^{k}/k!$. $e^{-2.5} \approx 0.0821$.
$\Pr(X = 0) \approx 0.0821$, $\Pr(X = 1) \approx 0.205$,
$\Pr(X = 2) \approx 0.257$. So $\Pr(X \leq 2) \approx 0.544$ and
$\Pr(X \geq 3) \approx 0.456$. Markov at $a = 5$:
$\Pr(X \geq 5) \leq \E[X]/5 = 0.5$. The exact value
$\Pr(X \geq 5) \approx 0.108$, so the Markov bound is loose by a
factor of about five.

\begin{exercise}
$X \sim \mathrm{Exp}(\lambda)$ with $\lambda = 0.01\,\text{h}^{-1}$.
Compute the mean lifetime, the median lifetime, and the probability
that the unit survives $200\,\text{h}$.
\end{exercise}

\paragraph{Solution.}
The mean of an exponential is the reciprocal of the rate,
$\E[X] = 1/\lambda = 1/0.01 = 100\,\text{h}$. The median solves
$F(m) = 1 - e^{-\lambda m} = 0.5$, so $e^{-\lambda m} = 0.5$ and
$m = \ln 2 / \lambda = 0.6931/0.01 \approx 69.3\,\text{h}$. The
median is below the mean, the signature right skew of the
exponential. Survival past $200\,\text{h}$ uses the complementary
CDF directly: $\Pr(X > 200) = e^{-\lambda \cdot 200} = e^{-2}
\approx 0.1353$. About one unit in seven survives two mean
lifetimes, and by memorylessness a unit that has already reached
$200\,\text{h}$ has the same residual-life distribution as a fresh
one.

\begin{exercise}
$X \sim \mathcal{N}(50, 16)$. Compute $\Pr(X > 55)$, $\Pr(X < 42)$,
and $\Pr(45 < X < 55)$ using the standard normal CDF $\Phi$.
\end{exercise}

\paragraph{Solution.}
The variance is $16$, so $\sigma = 4$. Standardise each query by
$z = (x - \mu)/\sigma$ with $\mu = 50$, $\sigma = 4$.
\begin{itemize}[leftmargin=*]
  \item $\Pr(X > 55)$: $z = (55 - 50)/4 = 1.25$, so
    $\Pr = 1 - \Phi(1.25) \approx 1 - 0.8944 = 0.1056$.
  \item $\Pr(X < 42)$: $z = (42 - 50)/4 = -2.0$, so
    $\Pr = \Phi(-2.0) \approx 0.0228$.
  \item $\Pr(45 < X < 55)$: the limits are $z = -1.25$ and
    $z = +1.25$, so $\Pr = \Phi(1.25) - \Phi(-1.25) = 0.8944 -
    0.1056 = 0.7888$.
\end{itemize}
The third answer is the central interval $\mu \pm 1.25\sigma$, which
captures about $79\,\%$ of the mass, consistent with the rule that
$\mu \pm 1\sigma$ holds about $68\,\%$ and $\mu \pm 2\sigma$ about
$95\,\%$.

\begin{exercise}
$(X, Y)$ has joint PMF given by the $3 \times 3$ table of the
reader's choice (with at least three non-zero entries). Compute
the marginals, the covariance, and the correlation.
\end{exercise}

\paragraph{Solution sketch.}
A worked example: $X, Y \in \{1, 2, 3\}$ with $p(1,1) = p(2,2) =
p(3,3) = 1/4$ and $p(1,3) = 1/4$; other cells zero. Marginals:
$p_{X}(1) = 1/2$, $p_{X}(2) = 1/4$, $p_{X}(3) = 1/4$; $p_{Y}(1) =
1/4$, $p_{Y}(2) = 1/4$, $p_{Y}(3) = 1/2$. $\E[X] = 1.75$, $\E[Y] =
2.25$. $\E[XY] = 1 \cdot 1 \cdot 1/4 + 4 \cdot 1/4 + 9 \cdot 1/4 +
3 \cdot 1/4 = 17/4 = 4.25$. $\Cov(X, Y) = 4.25 - 1.75 \cdot 2.25 =
0.3125$. Variances: $\Var(X) = \E[X^{2}] - (\E X)^{2} = (1/2 + 1 +
9/4) - 3.0625 = 0.6875$; $\Var(Y) = 0.6875$. Correlation
$0.3125/0.6875 \approx 0.455$. Any reader-chosen table with at
least three non-zero entries follows the same procedure.

\begin{exercise}
A sample of $n = 64$ observations has population mean
$\mu = 10$ and standard deviation $\sigma = 2$. Use the CLT to
estimate $\Pr(\bar X_{n} > 10.5)$.
\end{exercise}

\paragraph{Solution sketch.}
Standard error $\sigma/\sqrt{n} = 2/8 = 0.25$. $z = (10.5 - 10)/
0.25 = 2.0$. $\Pr(\bar X_{n} > 10.5) \approx 1 - \Phi(2.0) \approx
0.0228$.

\begin{exercise}
A two-state Markov chain has transition matrix
$\mat{P} = \begin{pmatrix} 0.8 & 0.2 \\ 0.4 & 0.6 \end{pmatrix}$.
Compute the stationary distribution $\vect{\pi}$ as the left
eigenvector of $\mat{P}$ for eigenvalue $1$. Verify $\vect{\pi}
\mat{P} = \vect{\pi}$ by direct multiplication.
\end{exercise}

\paragraph{Solution sketch.}
Stationarity: $\pi_{1} = 0.8 \pi_{1} + 0.4 \pi_{2}$, so $0.2 \pi_{1}
= 0.4 \pi_{2}$, i.e. $\pi_{1} = 2 \pi_{2}$. With $\pi_{1} + \pi_{2}
= 1$: $\pi_{2} = 1/3$, $\pi_{1} = 2/3$. Verify:
$(2/3, 1/3)\mat{P} = (0.8 \cdot 2/3 + 0.4 \cdot 1/3,\ 0.2 \cdot 2/3
+ 0.6 \cdot 1/3) = (0.5333 + 0.1333,\ 0.1333 + 0.2) = (2/3, 1/3)$.

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive each of the six consequences of the Kolmogorov axioms listed
in section 13.1: $\Pr(\emptyset) = 0$, $\Pr(A^{c}) = 1 - \Pr(A)$,
monotonicity, inclusion-exclusion for two events, continuity from
below, and Boole's inequality.
\end{exercise}

\paragraph{Solution sketch.}
$\Pr(\emptyset) = 0$: $\Omega = \Omega \cup \emptyset$, disjoint, so
$1 = 1 + \Pr(\emptyset)$. $\Pr(A^{c}) = 1 - \Pr(A)$: $\Omega = A
\cup A^{c}$, disjoint, gives $1 = \Pr(A) + \Pr(A^{c})$.
Monotonicity ($A \subseteq B$): $B = A \cup (B \setminus A)$,
disjoint, gives $\Pr(B) = \Pr(A) + \Pr(B \setminus A) \geq \Pr(A)$.
Inclusion-exclusion: $A \cup B = A \cup (B \setminus A)$ with $A$
and $B \setminus A$ disjoint; $B = (A \cap B) \cup (B \setminus A)$
disjoint; combining gives the formula. Continuity from below: let
$B_{1} = A_{1}$, $B_{i} = A_{i} \setminus A_{i-1}$, the $B_{i}$ are
pairwise disjoint with $\bigcup B_{i} = \bigcup A_{i}$ and
$\bigcup_{i \leq n} B_{i} = A_{n}$. Countable additivity gives the
result. Boole: $A_{1} \cup \cdots \cup A_{n}$ written as a disjoint
union $A_{1} \cup (A_{2} \setminus A_{1}) \cup \cdots$ with each
piece a subset of the original $A_{i}$; sum and use monotonicity.

\begin{exercise}
Derive Bayes' rule from the multiplication rule applied two ways.
Show that the posterior depends on the prior even when the
likelihood ratio is large.
\end{exercise}

\paragraph{Solution sketch.}
Multiplication rule: $\Pr(A \cap B) = \Pr(A \mid B) \Pr(B) = \Pr(B
\mid A) \Pr(A)$. Solve for $\Pr(B \mid A)$: $\Pr(B \mid A) = \Pr(A
\mid B) \Pr(B) / \Pr(A)$. Replace $\Pr(A)$ by the law of total
probability over a partition $\{B_{j}\}$ to obtain the full Bayes
form. With $\Pr(A \mid B)/\Pr(A \mid B^{c}) = 100$ (likelihood
ratio) and $\Pr(B) = 10^{-4}$, the posterior odds are $100 \cdot
(10^{-4}/(1-10^{-4})) \approx 10^{-2}$, giving posterior probability
$\approx 0.0099$. A large likelihood ratio multiplied by a small
prior is still small.

\begin{exercise}
Construct three events $A_{1}, A_{2}, A_{3}$ on a fair-coin sample
space that are pairwise independent but not mutually independent.
Verify both properties by direct computation.
\end{exercise}

\paragraph{Solution sketch.}
Toss a fair coin twice, $\Omega = \{HH, HT, TH, TT\}$ with the
uniform measure. Let $A_{1} = \{HH, HT\}$ (first toss heads),
$A_{2} = \{HH, TH\}$ (second toss heads), $A_{3} = \{HH, TT\}$
(matching tosses). Each event has probability $1/2$. Pairwise:
$A_{1} \cap A_{2} = A_{1} \cap A_{3} = A_{2} \cap A_{3} = \{HH\}$
with probability $1/4 = (1/2)(1/2)$, so all pairs are independent.
Joint: $A_{1} \cap A_{2} \cap A_{3} = \{HH\}$ with probability
$1/4$, but the product $(1/2)^{3} = 1/8 \neq 1/4$.

\begin{exercise}
For $X \sim \mathrm{Geom}(p)$, derive the memoryless property
$\Pr(X > m + n \mid X > m) = \Pr(X > n)$.
\end{exercise}

\paragraph{Solution sketch.}
$\Pr(X > k) = (1 - p)^{k}$ since the first $k$ trials must all fail.
Conditional: $\Pr(X > m + n \mid X > m) = \Pr(X > m + n) / \Pr(X >
m) = (1-p)^{m+n} / (1-p)^{m} = (1-p)^{n} = \Pr(X > n)$. The chance
of going at least $n$ further steps does not depend on how far the
process has already gone.

\begin{exercise}
Show that $\Var(aX + bY) = a^{2}\Var(X) + b^{2}\Var(Y) + 2ab
\Cov(X, Y)$. Specialise to the case $a = b = 1$ and to the case
of independent $X, Y$.
\end{exercise}

\paragraph{Solution sketch.}
Let $U = aX + bY$. Then $U - \E U = a(X - \E X) + b(Y - \E Y)$.
$\Var(U) = \E[(U - \E U)^{2}] = a^{2}\E[(X - \E X)^{2}] + b^{2}
\E[(Y - \E Y)^{2}] + 2ab \E[(X - \E X)(Y - \E Y)]$, which gives the
formula. With $a = b = 1$: $\Var(X + Y) = \Var X + \Var Y + 2\Cov(X,
Y)$. With $X, Y$ independent: $\Cov = 0$, and the variance of the
sum reduces to the sum of variances.

\begin{exercise}
Prove the Markov inequality from the definition of expectation.
Then derive Chebyshev's inequality as a corollary applied to
$(X - \mu)^{2}$.
\end{exercise}

\paragraph{Solution sketch.}
For $X \geq 0$ and $a > 0$: $a \cdot \mathbf{1}\{X \geq a\} \leq X$
(pointwise). Take expectations: $a \Pr(X \geq a) \leq \E[X]$. Divide
by $a$ to obtain Markov. Chebyshev: apply Markov to $(X - \mu)^{2}
\geq 0$ at threshold $k^{2}\sigma^{2}$: $\Pr((X - \mu)^{2} \geq
k^{2}\sigma^{2}) \leq \E[(X - \mu)^{2}]/(k^{2}\sigma^{2}) = 1/k^{2}$.
The event $(X - \mu)^{2} \geq k^{2}\sigma^{2}$ is identical to
$|X - \mu| \geq k\sigma$.

\begin{exercise}
Show that $X \sim \mathrm{Bin}(n, p)$ has $\E[X] = np$ and
$\Var(X) = np(1 - p)$ by writing $X = \sum_{i=1}^{n} Y_{i}$ with
$Y_{i} \sim \mathrm{Bern}(p)$ independent and using linearity of
expectation and the variance formula for sums.
\end{exercise}

\paragraph{Solution sketch.}
$Y_{i} \in \{0, 1\}$ with $\E[Y_{i}] = p$, $\E[Y_{i}^{2}] = p$, so
$\Var(Y_{i}) = p - p^{2} = p(1 - p)$. Linearity gives $\E[X] =
\sum_{i} \E[Y_{i}] = np$. Independence kills covariances:
$\Var(X) = \sum_{i} \Var(Y_{i}) = np(1 - p)$.

\begin{exercise}
For the bivariate Gaussian with correlation $\rho$, show that the
conditional distribution of $Y$ given $X = x$ is Gaussian with mean
$\mu_{Y} + \rho (\sigma_{Y}/\sigma_{X})(x - \mu_{X})$ and variance
$\sigma_{Y}^{2}(1 - \rho^{2})$. Identify the regression line and
the residual variance in this expression.
\end{exercise}

\paragraph{Solution sketch.}
Write the joint density as the product $f_{Y \mid X}(y \mid x)
f_{X}(x)$. The exponent of the joint Gaussian is a quadratic form
in $(x - \mu_{X}, y - \mu_{Y})$; completing the square in $y$ at
fixed $x$ gives a quadratic in $y$ with apex at $\mu_{Y} +
\rho(\sigma_{Y}/\sigma_{X})(x - \mu_{X})$ and coefficient
$1/(2\sigma_{Y}^{2}(1 - \rho^{2}))$. The conditional density is
therefore Gaussian with the stated mean and variance. The
conditional mean as a function of $x$ is the regression line; the
conditional variance $\sigma_{Y}^{2}(1 - \rho^{2})$ is the residual
variance. At $\rho = \pm 1$ the residual variance is zero; at
$\rho = 0$ it equals the marginal $\sigma_{Y}^{2}$.

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
A coin is flipped $1000$ times and lands heads $530$ times.
Estimate, before computing, the standard deviation of the number
of heads under the null hypothesis of fairness. Compare your
estimate to the formula $\sqrt{np(1-p)}$.
\end{exercise}

\paragraph{Solution sketch.}
First-pass estimate: $\sqrt{n} = \sqrt{1000} \approx 32$, times
$\sqrt{p(1-p)} = 0.5$, gives $\approx 16$. Formula: $\sqrt{1000
\cdot 0.5 \cdot 0.5} = \sqrt{250} \approx 15.8$. The observed $530$
is $(530 - 500)/15.8 \approx 1.9$ standard deviations above the
mean, marginal-significant but not strong evidence against
fairness.

\begin{exercise}
A network interface receives, on average, $50$ packets per second.
Estimate the probability of receiving at least $70$ packets in one
second under a Poisson model. Compare to the Gaussian approximation
$\mathcal{N}(50, 50)$.
\end{exercise}

\paragraph{Solution sketch.}
$\mathrm{Pois}(50)$ has mean $50$ and standard deviation
$\sqrt{50} \approx 7.07$. Seventy is $(70 - 50)/7.07 \approx 2.83$
standard deviations above the mean. Gaussian approximation:
$\Pr(K \geq 70) \approx 1 - \Phi(2.83 - 0.071) \approx 1 -
\Phi(2.76) \approx 0.003$, applying a continuity correction of
$0.5$ in the standardised units. (Exact Poisson tail computed via
the cumulative formula gives $\approx 0.0033$, in close agreement.)

\begin{exercise}
Five components are arranged in series, each with reliability
$0.99$. Estimate the system reliability to one significant figure
without using a calculator. Then compute the exact value.
\end{exercise}

\paragraph{Solution sketch.}
First-order estimate: failure probability about $5 \cdot 0.01 =
0.05$, so reliability about $0.95$ (one significant figure). Exact:
$0.99^{5} = (1 - 0.01)^{5} \approx 1 - 0.05 + 0.001 - \ldots
\approx 0.951$.

\begin{exercise}
The same five components are arranged in parallel (the system
fails only when all five fail) with the same individual
reliability. Estimate the system unreliability and the system
reliability without a calculator. Compute and compare.
\end{exercise}

\paragraph{Solution sketch.}
Each component has failure probability $0.01$; under independence
all five fail with probability $0.01^{5} = 10^{-10}$. System
reliability $1 - 10^{-10}$. The estimate without a calculator: every
factor of $10$ in the exponent corresponds to roughly $33\,\%$
shrinkage in base ten, so $10^{-10}$ is sixteen powers of two below
$10^{-5}$. The contrast with series ($1 - 0.95 = 0.05$ failure
probability) is nine orders of magnitude.

\begin{exercise}
A diagnostic test has $99\,\%$ sensitivity, $95\,\%$ specificity,
and is applied to a population with prevalence $0.001$. Estimate
the posterior probability of disease given a positive test before
computing. Then compute and compare. Identify the dominant term in
the denominator.
\end{exercise}

\paragraph{Solution sketch.}
Estimate: false-positive rate $0.05$, prevalence $0.001$, so the
diseased population contributes about $0.001 \cdot 0.99 \approx
10^{-3}$ to the positive count while the healthy population
contributes $0.999 \cdot 0.05 \approx 0.05$. Posterior odds about
$10^{-3} / 0.05 = 0.02$, so posterior probability about $0.02$, or
$2\,\%$. Exact: $0.99 \cdot 0.001 / (0.99 \cdot 0.001 + 0.05 \cdot
0.999) = 0.00099 / 0.05094 \approx 0.0194$. Dominant denominator
term is the false-positive contribution $0.05 \cdot 0.999$, which
exceeds the true-positive contribution by a factor of fifty.

\begin{exercise}
A queue receives customers at average rate $\lambda = 5$ per hour
under a Poisson model. Estimate the median inter-arrival time and
the probability that the next inter-arrival exceeds 30 minutes.
\end{exercise}

\paragraph{Solution sketch.}
Inter-arrivals are $\mathrm{Exp}(5\,\text{h}^{-1})$. Median $\ln 2 /
\lambda = 0.693 / 5\,\text{h} \approx 8.3\,\text{min}$. $\Pr(\text{
next} > 0.5\,\text{h}) = e^{-5 \cdot 0.5} = e^{-2.5} \approx 0.082$.

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate $10^{4}$ throws of a fair die. Plot the empirical
distribution; verify it is uniform within sampling error. Repeat
with $10^{2}$ and $10^{6}$ throws and comment on the convergence
rate.
\end{exercise}

\paragraph{Solution sketch.}
Expected count per face at $10^{4}$ throws is $\approx 1667$ with
standard deviation $\sqrt{1667 \cdot 5/6} \approx 37$, so each bar
should be within about $\pm 75$ ($\pm 2\sigma$) of the expectation.
At $10^{2}$ throws the standard deviation per bar is about $\sqrt{
16.7 \cdot 5/6} \approx 3.7$, swamping the spacing. At $10^{6}$
throws the standard deviation is about $370$ on an expectation of
$166\,667$, giving relative error below $0.3\,\%$. The
$1/\sqrt{N}$ scaling is visible in the contrast across the three
counts.

\begin{exercise}
Simulate $10^{4}$ realisations of $\bar X_{n}$ for $X_{i}
\stackrel{iid}{\sim} \mathrm{Exp}(1)$ at $n = 5$, $n = 30$, and
$n = 100$. Plot a histogram of each and overlay the CLT Gaussian
approximation $\mathcal{N}(1, 1/n)$. Comment on the rate at which
the Gaussian fit improves.
\end{exercise}

\paragraph{Solution sketch.}
Use \repopath{volumes/02-form/ch13-probability/code/clt_demo.py}.
Sample skewness of the standardised mean should be roughly $2/
\sqrt{n}$ for $\mathrm{Exp}(1)$ (Berry-Esseen-style): $\approx
0.89, 0.37, 0.20$ for $n = 5, 30, 100$. At $n = 5$ the histogram
still leans visibly right; at $n = 30$ the asymmetry is faint; at
$n = 100$ the overlay is indistinguishable from the data at the
precision of the plot.

\begin{exercise}
Simulate $10^{4}$ realisations of the maximum of $n$ iid
$\mathrm{Exp}(1)$ random variables for $n = 10, 100, 1000$. Compare
the empirical distribution to the asymptotic Gumbel limit (look up
the form). Discuss why the CLT does not apply here.
\end{exercise}

\paragraph{Solution sketch.}
The maximum $M_{n}$ of iid $\mathrm{Exp}(1)$ satisfies $M_{n} -
\ln n \to_{D} G$, where $G$ has the standard Gumbel CDF $\exp(-
e^{-x})$. Empirical means should be approximately $\ln 10 \approx
2.30$, $\ln 100 \approx 4.61$, $\ln 1000 \approx 6.91$. The CLT
does not apply because the maximum is not a sum: the maximum is
dominated by a single tail observation and converges to an
extreme-value distribution, not to a Gaussian. Volume X (failure
and reliability) returns to the extremal limit laws.

\begin{exercise}
Simulate the two-state weather chain of section 13.7 for $10^{5}$
steps. Plot the running fraction of sunny days and compare to the
theoretical $5/6$.
\end{exercise}

\paragraph{Solution sketch.}
Use \repopath{volumes/02-form/ch13-probability/code/markov_chain_simulation.py}
with \texttt{-{}-steps 100000}. The running fraction settles within
about $1\,\%$ of $5/6 \approx 0.833$ by $10^{4}$ steps and within
$0.3\,\%$ by $10^{5}$ steps, consistent with the $1/\sqrt{t}$ rate
modulated by the chain's spectral gap.

\begin{exercise}
Implement a simple random walk on $\{0, 1, \ldots, 100\}$ with
reflecting boundaries. Run for $10^{6}$ steps. Compare the
empirical occupancy to the uniform stationary distribution. Estimate
the mixing time.
\end{exercise}

\paragraph{Solution sketch.}
Reflecting random walk: at $0$, deterministically move to $1$; at
$100$, deterministically move to $99$; elsewhere, move $\pm 1$ with
probability $1/2$. The chain is irreducible but periodic with
period $2$; apply a lazy correction (stay with probability $1/2$,
otherwise move) to break the period. Stationary distribution is
uniform on $\{0, \ldots, 100\}$ (each interior state has equal
balanced flow). Mixing time scales like $k^{2}$ for $k = 100$,
giving order $10^{4}$ steps; the running occupancy should be within
$0.01$ in total variation of uniform by $10^{5}$ steps.

\begin{exercise}
Simulate $n = 50$ correlated Bernoulli variables (use a Gaussian
copula or a hidden common factor of the reader's design). Compare
the empirical variance of the sum to the variance under the
independence assumption $np(1-p)$. Quantify the size of the
correlation effect.
\end{exercise}

\paragraph{Solution sketch.}
Use \repopath{volumes/02-form/ch13-probability/code/correlated_bernoulli.py}.
At $p = 0.10$, $n = 50$, $\rho = 0.0$ the empirical variance is
$\approx np(1-p) = 4.5$. At $\rho = 0.3$ it rises to about $20$;
at $\rho = 0.7$ to about $50$; at $\rho = 0.9$ to about $80$. The
ratio empirical-to-independent grows roughly linearly in $\rho$ at
moderate $\rho$; the asymptote at $\rho = 1$ is $n^{2}p(1-p)/[np
(1-p)] = n$.

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a two-stage screening test for a manufacturing defect with
prior $0.5\,\%$. The first stage has $99\,\%$ sensitivity and
$90\,\%$ specificity; the second stage has $95\,\%$ sensitivity and
$99\,\%$ specificity. Compute the post-second-test posterior given
two positives. Discuss whether the second test should be
conditionally independent of the first given the unit's true
state, and what design steps support that.
\end{exercise}

\paragraph{Discussion.}
First stage: $\Pr(D \mid +_{1}) = 0.99 \cdot 0.005 / (0.99 \cdot
0.005 + 0.10 \cdot 0.995) = 0.00495 / 0.10445 \approx 0.0474$. With
the updated prior $0.0474$, second stage: $\Pr(D \mid +_{1}, +_{2})
= 0.95 \cdot 0.0474 / (0.95 \cdot 0.0474 + 0.01 \cdot 0.9526) =
0.04503 / 0.05456 \approx 0.825$. The two-stage posterior is above
$80\,\%$, an order of magnitude tighter than either stage alone.
The chain of updates assumes conditional independence given the
unit's true state: the second test must not share a failure mode
with the first. Design supports: use a physically different
measurement principle (optical for stage one, electrical for stage
two), batch the two tests at different times of day, calibrate them
against independent references. A pair of tests that share an
operator, a calibration standard, or an environmental sensitivity
will have correlated errors and the apparent gain over a single
test will overstate the real gain.

\begin{exercise}
Design a Monte Carlo estimator of $\pi$ that samples points uniformly
in $[-1, 1]^{2}$ and counts the fraction inside the unit disk.
Compute the variance of the estimator, and the sample size required
to achieve a $95\,\%$ confidence interval of half-width $0.001$.
\end{exercise}

\paragraph{Discussion.}
The indicator $\mathbf{1}\{X^{2} + Y^{2} \leq 1\}$ is Bernoulli
with $p = \pi/4 \approx 0.7854$. Estimator $\hat\pi = 4 \hat p$
with $\hat p$ the sample mean. Variance of $\hat p$ is $p(1 -
p)/N \approx 0.1685/N$; variance of $\hat\pi$ is
$16 \cdot 0.1685/N \approx 2.70/N$. Half-width
$1.96 \sqrt{2.70/N} = 0.001$ gives $N = 1.96^{2} \cdot 2.70 / 10^{-6}
\approx 1.04 \times 10^{7}$. Roughly ten million samples for three
significant figures; quasi-Monte Carlo or stratified sampling
would reduce the required sample size by orders of magnitude and
should be considered for production use.

\begin{exercise}
Design a redundancy architecture for a safety-critical sensor that
must achieve a failure probability below $10^{-6}$ over a one-year
mission. Single-sensor failure probability is $10^{-3}$. Document
the dependence channels between any two sensors and propose
mitigations that reduce the joint-failure probability below the
product of marginals.
\end{exercise}

\paragraph{Discussion.}
Under independence, two sensors give joint failure probability
$10^{-6}$, just at the requirement; three sensors give $10^{-9}$,
comfortably below. The independence assumption is almost never
warranted off the page. Dependence channels in practice: shared
power supply, shared signal conditioning electronics, shared
firmware version (a software-defect failure is correlated across
identical units), shared environmental conditions (temperature,
vibration, EMI), shared calibration date (drift drift in parallel
across units calibrated together), shared operator. Mitigations
that lower the joint probability include diverse hardware
(different sensor technologies measuring the same quantity, e.g.
optical and resistive thermometers), diverse software (independent
implementations, ideally by separate teams), independent power
domains, physical separation against environmental causes, staggered
calibration schedules. The honest design target is to budget the
correlation explicitly: assume a residual joint failure probability
$10^{-5}$ or so after diversification, document the channels that
remain, and add a third unit if the remainder is still above
threshold.

\begin{exercise}
Design an A/B test for two variants of a product feature, with a
target effect size of one percentage point change in a click-rate
of nominal value $5\,\%$, $\alpha = 0.05$, and power $0.80$.
Compute the required sample size per group. Document at least one
dependence between users that would inflate the effective Type I
error rate above $\alpha$ if ignored.
\end{exercise}

\paragraph{Discussion.}
For a two-proportion $z$-test with $p = 0.05$, $\Delta = 0.01$, two
sided $\alpha = 0.05$ ($z_{0.975} = 1.96$), power $0.80$
($z_{0.80} = 0.84$): $n \approx (z_{\alpha/2} + z_{\beta})^{2}
\cdot 2 p(1 - p)/\Delta^{2} = (2.80)^{2} \cdot 2 \cdot 0.0475 /
0.0001 \approx 7\,440$ per group. Dependence channels that inflate
the Type I rate if ignored: clustering (multiple sessions per
user counted as independent observations), shared device or
network conditions, cohort effects (the same news event affects
both arms differently), bot traffic correlated across the
randomisation, and time-of-day or day-of-week effects when the
randomisation is not balanced over time.

\begin{exercise}
Design a queueing model for a service desk that receives customers
at a Poisson rate $\lambda$ and serves them with exponential
service time of rate $\mu > \lambda$. Compute the stationary
distribution of the number in queue. Discuss what happens when the
service-time distribution is heavy-tailed instead.
\end{exercise}

\paragraph{Discussion.}
The model is M/M/1. Utilisation $\rho = \lambda/\mu < 1$. The
stationary distribution of the number in system $N$ is geometric:
$\Pr(N = k) = (1 - \rho) \rho^{k}$ for $k = 0, 1, 2, \ldots$ Mean
$\E[N] = \rho/(1 - \rho)$, which grows without bound as $\rho \to
1$. Mean waiting time (Little's law) is $\E[N]/\lambda$. A numeric
anchor: at $\lambda = 8$ per hour and $\mu = 10$ per hour,
$\rho = 0.8$, so $\E[N] = 0.8/0.2 = 4$ customers in system and a
mean time in system of $4/8 = 0.5\,\text{h}$. Pushing utilisation to
$\rho = 0.9$ by raising $\lambda$ to $9$ gives $\E[N] = 9$ and a
mean time of $1.0\,\text{h}$: a $12.5\,\%$ increase in load more than
doubles the wait, the hallmark non-linearity of the
$\rho/(1 - \rho)$ term near saturation. With a
heavy-tailed (Pareto-like) service-time distribution, the
Pollaczek-Khinchine formula applies for M/G/1 and the mean wait
gains a $1 + C_{s}^{2}$ multiplier where $C_{s}$ is the coefficient
of variation of service times; for distributions with very large or
infinite variance the mean wait can be much larger than the M/M/1
prediction at the same $\rho$, and tail-quantile waits (the $99$th
percentile) can be larger still. Volume IX (control) develops the
machinery; the engineering takeaway is that heavy service-time
tails make queues much worse than utilisation alone would suggest.

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A safety analyst reports a joint failure probability of $10^{-9}$
for a triple-redundant system with individual channel failure
probabilities of $10^{-3}$. State the implicit independence
assumption and identify three candidate dependence channels that
the report does not address. Estimate, with one paragraph of
reasoning, what the joint failure probability would be under
moderate dependence ($\rho = 0.3$ between channel failures).
\end{exercise}

\paragraph{Discussion.}
The $10^{-9}$ figure assumes the three channel failures are
mutually independent: $(10^{-3})^{3}$. Candidate dependence
channels: shared power supply (a single failure depowers all three),
shared firmware (a software defect affects every channel identically),
shared environmental envelope (high temperature or vibration affects
all three at once), shared calibration batch (drift correlated across
the channels). At $\rho = 0.3$ between channel failures the joint
probability is not simply the product; a tractable model takes the
correlation through a hidden common factor that drives a fraction
$\rho$ of the failure variance. The Bahadur-form upper bound for
pairwise positively correlated indicators gives, for small marginals
$p$ and pairwise correlation $\rho$, a joint probability of order
$p \cdot \max(p, \rho)^{n - 1}$; at $p = 10^{-3}$ and $\rho = 0.3$
this is $10^{-3} \cdot 0.09 \approx 10^{-4}$, five orders of
magnitude above the independence figure. The honest report would
state the independence assumption explicitly and present a range.

\begin{exercise}
A team reports that two variables in their dataset have Pearson
correlation $0.02$ and concludes the variables are independent. Read
the conclusion. Identify what claim the data support and what claim
they do not, and propose a test that would speak to independence
rather than to absence of linear association.
\end{exercise}

\paragraph{Discussion.}
The data support the claim that the linear component of dependence
is small. They do not support independence: the $Y = X^{2}$
counterexample of section 13.5 has $\rho = 0$ and total dependence.
Tests that speak to independence rather than to linear association:
distance correlation (Sz{\'e}kely-Rizzo), maximal information
coefficient (MIC), conditional mutual information, kernel-based
independence tests (HSIC), or a non-parametric permutation test
combined with a non-linear test statistic such as the rank
correlation of $X$ with $|Y|$. A scatter plot is the cheapest and
often the most informative check: a U-shape, a wedge, or a clear
non-linear pattern overrides any Pearson coefficient.

\begin{exercise}
A reader fits a Gaussian distribution to the daily returns of an
asset and computes $99\,\%$ confidence bounds on weekly portfolio
losses using $\sigma\sqrt{n}$ scaling. Read the procedure. State
the distributional assumption the procedure depends on and the
empirical features of asset returns that violate it. Estimate
qualitatively whether the procedure overstates or understates tail
risk.
\end{exercise}

\paragraph{Discussion.}
The procedure assumes daily returns are iid Gaussian; weekly
returns then are also Gaussian with standard deviation
$\sigma\sqrt{5}$. Empirical features that violate the assumption:
returns are heavy-tailed (excess kurtosis above the Gaussian's
$3$), have volatility clustering (variance is autocorrelated, an
ARCH/GARCH feature), and are skewed (large negative moves are more
likely than the Gaussian predicts). All three violations make the
true tail thicker than the Gaussian extrapolation; the procedure
\emph{understates} tail risk, often by an order of magnitude at
the $99\,\%$ quantile and by more at the $99.9\,\%$ quantile. The
2008 crisis case in section 13.9 is one episode in a long history
of this pattern.

\begin{exercise}
A reliability engineer observes that of $1000$ units that have run
for $1\,\text{yr}$, $5$ have failed. They conclude that the failure
probability over the next year is $5/1000 = 0.005$. Read the
inference. State at least three conditions under which the
inference is valid and at least three under which it is not.
\end{exercise}

\paragraph{Discussion.}
Conditions under which $\hat p = 5/1000$ is a defensible estimate
for the next year's failure probability: the units were a random
sample of the population the next year will draw from; the failure
mechanism has constant hazard, so past-year and next-year
probabilities are exchangeable; operating conditions, supplier,
materials, and software version are unchanged between the two years.
Conditions under which the inference fails: the hazard rises with
age (an aging mechanism like wear or fatigue), so next year's $p$
is larger; the hazard falls with age (infant mortality), so next
year's $p$ is smaller; the sample is biased (the failed units were
removed and the survivors are not representative of the next
year's units); a design or supplier change is planned; the failure
count of $5$ has a Poisson uncertainty of $\sqrt{5} \approx 2.2$,
so the estimate carries a $95\,\%$ confidence interval of roughly
$[0.001, 0.012]$, an order of magnitude wide.

\begin{exercise}
Read section 13.9's claim that the union bound is a conservative
substitute for an independence-based joint failure calculation.
Identify the regime in which the two are nearly equal and the
regime in which they differ by orders of magnitude. State the
quantity an engineer would need to know to choose between them
honestly.
\end{exercise}

\paragraph{Discussion.}
Independence: $\Pr(\bigcup A_{i}) = 1 - \prod (1 - p_{i})$. Union
bound: $\Pr(\bigcup A_{i}) \leq \sum p_{i}$. For small $p_{i}$ with
$\sum p_{i} \ll 1$, the two agree to first order: $1 - (1 - p)^{n}
\approx np$ when $np \ll 1$. They differ when $np$ approaches $1$:
union bound saturates at $1$ while independence remains below it.
For large positive correlation the joint event probability falls
toward a single marginal $p$, far below either of the two
estimates above. The quantity the engineer needs is the
correlation structure of the events, summarised at minimum by the
pairwise correlation matrix and a structural model of any shared
upstream cause. Without that, the union bound is the safer reach;
the independence calculation is reportable only after explicit
justification that the events do not share a generative mechanism.
The companion script
\repopath{volumes/02-form/ch13-probability/code/union_bound_check.py}
tabulates the gap explicitly.

\begin{exercise}
A model assigns a posterior probability of $0.95$ to a hypothesis
$H$ on the basis of one positive test result. The test has
sensitivity $0.99$ and specificity $0.99$. Read the model. Use
Bayes' rule to back out the prior the model used, and assess
whether that prior is plausible given the engineering context the
reader brings.
\end{exercise}

\paragraph{Discussion.}
Posterior $0.95 = 0.99 \pi / (0.99 \pi + 0.01 (1 - \pi))$. Cross
multiply: $0.95 \cdot 0.99 \pi + 0.95 \cdot 0.01 (1 - \pi) = 0.99
\pi$, so $0.0095 (1 - \pi) = 0.99 \pi - 0.9405 \pi = 0.0495 \pi$.
Hence $0.0095 = (0.0095 + 0.0495) \pi$, giving $\pi \approx 0.161$.
The implicit prior is about $16\,\%$, far higher than the rare-event
priors of $10^{-3}$ or smaller that motivate most safety-critical
diagnostic tests. The reader should ask whether a $16\,\%$ prior
matches the actual base rate of the condition the test is meant to
detect; in rare-event applications it almost certainly does not, and
the model's posterior is consequently miscalibrated. The base-rate
fallacy of section 13.2 is the failure mode.
