\chapter{Heat exchangers and thermal systems}
\label{vol04:ch06}

\epigraph{The function of a heat exchanger is to transfer heat from one
fluid stream to another; its design problem is to do so with the
smallest area, the smallest pressure drop, and the fewest tube
failures.}{Donald Q.\ Kern, \emph{Process Heat Transfer} (1950),
preface \cite[p.~v]{text:kern-process-heat}.}

\chapmeta{Half-life: durable. Archetypes: transport, optimisation, failure. Exercise target: 28. Project track: design plus software verification of a specified-duty heat exchanger. Hazard class: low (paperwork and software). Reader path default: standard, with sections explicitly tagged. Named cases: none required in this chapter; tube-leak and fouling failures are described by mechanism.}

\noindent
A heat exchanger is a device that transfers heat between two fluid
streams without mixing them. The design problem is to identify the
geometry, the materials, and the operating conditions that deliver
the required duty (heat transferred per unit time) at acceptable
cost, pressure drop, fouling, and corrosion. The chapter develops the
canonical heat-exchanger types, the two working design methods
(LMTD and NTU-effectiveness), the pumping-power and storage
considerations that turn the heat-exchanger into part of a thermal
system, and the failure modes that consume the bulk of operating
maintenance.

\section{Heat exchanger types}\pathtag{core}

\engterm{Shell-and-tube exchangers} are the workhorse of industrial
heat transfer. A bundle of tubes inside a cylindrical shell carries
one fluid; the other fluid flows through the shell, baffled to flow
across the tube bundle. Standard configurations:

\begin{itemize}[leftmargin=*]
  \item Fixed-tubesheet: simplest construction, no expansion
    accommodation; suitable when shell and tube fluids are at similar
    temperatures or when tube-bundle removal for cleaning is not
    required.
  \item U-tube: the tubes form a U-shape with both ends in the same
    tubesheet; allows thermal expansion but makes tube cleaning
    harder.
  \item Floating-head: one tubesheet is fixed and the other floats
    inside the shell, allowing both expansion and bundle removal;
    the most common industrial configuration for severe service.
\end{itemize}

The Tubular Exchanger Manufacturers Association (TEMA) standards
\cite{text:tema-standards} define the working dimensional and
mechanical-design conventions: tube pitch, baffle cut, shell-side
clearances, nozzle locations, and the three-letter codes (front-
head, shell, rear-head) that uniquely specify a configuration.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-shell-and-tube-schematic.tex}

\engterm{Plate-and-frame exchangers} consist of corrugated metal
plates clamped together with gaskets, forming alternating channels
for the two fluids. Plate-and-frame designs offer high effectiveness
(close approach between the two outlet temperatures), high heat-
transfer coefficients (turbulence induced by the corrugation), and
straightforward cleaning (the plates can be unbolted and washed). The
gaskets limit the operating temperature ($\sim 180\,\degC$ for nitrile
or EPDM) and pressure ($\sim 25\,\text{bar}$), restricting plate-and-
frame to moderate-condition services. Gasket-free plate exchangers
(welded plate pairs, brazed plate exchangers, plate-and-fin) extend
the operating envelope at the cost of cleanability.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-plate-exchanger.tex}

\engterm{Compact exchangers} use small hydraulic diameters and
extended-surface geometries to achieve high heat-transfer area per
unit volume. Examples: plate-fin exchangers (used in cryogenic
service and in aerospace thermal management), printed-circuit
exchangers (chemically etched flow channels in stacked plates, used
in supercritical-CO$_2$ and hydrogen service), and microchannel
exchangers (used in electronics cooling and small-scale heat pumps).

\engterm{Air-cooled exchangers (fin-fan coolers)} reject heat to
ambient air through finned tubes with forced draft from large fans.
Air-cooled exchangers avoid the water-supply and water-treatment
requirements of cooling-water systems but have lower heat-transfer
coefficients and larger footprints. They are the dominant choice in
remote installations (gas-processing plants in desert locations,
remote pipeline compressor stations) and a common choice for
condenser duty in dry climates.

\engterm{Direct-contact exchangers} mix the two fluids deliberately
to maximise heat transfer; cooling towers (water cooled by direct
contact with air), spray chambers (gas cooled by direct contact with
spray water), and quench drums all transfer heat by direct contact.
The constraint is that the two fluids must be compatible (the cooling
tower's losses to evaporation, drift, and blowdown are the price for
direct contact).

The choice between exchanger types is set by the service: thermal
duty, fluid properties, fouling tendency, materials of construction,
maintenance access, footprint, and capital cost. Process-industry
practice maintains design heuristics for each class; the working
engineer matches the heuristic to the specific service. A few rules
carry most cases. Shell-and-tube wins on pressure and temperature
envelope, on tolerance of dirty and fouling fluids, and on the
availability of code-stamped mechanical design, which is why it is the
default for high-pressure and high-temperature process duty. Plate-and-
frame wins when a close temperature approach is wanted (two outlet
temperatures within a few kelvin of crossing), when the fluids are clean
or the unit must be opened often for cleaning, and when floor space is
tight, which is why it dominates dairy, district-heating substation, and
HVAC service. Air-cooled units win where cooling water is scarce or
expensive to treat and where a larger footprint is acceptable. Compact
exchangers win where surface area per unit volume is the binding
constraint, in cryogenic, aerospace, and electronics service, at the cost
of small channels that foul readily and clean poorly. The first decision
in a design is therefore not the size but the class, because the class
fixes which trade-offs the later sizing can still touch.

\subsection{The compactness metric and where each class sits}

The word ``compact'' has a precise meaning worth pinning down, because it is the
axis along which the classes are ranked. The \engterm{area density} $\beta$ is the
heat-transfer surface area packed into a unit of exchanger volume, measured in
$\text{m}^2/\text{m}^3$, and it is set almost entirely by the hydraulic diameter of
the flow passages: a bundle of small channels packs far more wetted surface into a
given volume than a bundle of large ones, because surface scales as the passage
perimeter while volume scales as its area. A shell-and-tube unit with
$19\,\si{\milli\meter}$ tubes runs an area density near $70$ to $500\,\text{m}^2/
\text{m}^3$; a plate-and-frame unit reaches $200$ to $1000$; a plate-fin or
printed-circuit compact exchanger reaches $1000$ to $5000$; and the human lung, the
limiting biological case, packs about $20{,}000\,\text{m}^2/\text{m}^3$. The
conventional dividing line calls a surface compact above about $700\,\text{m}^2/
\text{m}^3$, which is the region small channels reach and round tubes do not. The
metric explains the class trade-offs at a stroke: small channels buy area density
but raise the pumping cost per unit flow and foul readily because a deposit that is
a nuisance in a $19\,\si{\milli\meter}$ tube blocks a $1\,\si{\milli\meter}$
channel, so compact surfaces are reserved for clean fluids where volume or weight is
the binding constraint. The area density also fixes the exchanger's response and its
inventory: a compact core holds little fluid and warms quickly, which suits a duty
that must track a changing load, while a shell-and-tube unit holds a large inventory
that buffers upsets but lags a setpoint. Reading a service against the area-density
axis is the quantitative form of the class choice: name the binding constraint,
volume, weight, cleanliness, pressure, or response, and the area density the service
can tolerate points at the class before a single dimension is fixed.

\section{LMTD method}\pathtag{core}

The \engterm{log-mean temperature difference} (LMTD) method is the
classical design approach: given the four terminal temperatures (hot
inlet, hot outlet, cold inlet, cold outlet) and the heat-transfer
rate, find the required heat-transfer area. The method is the
working tool when the heat duty is specified.

For a counter-flow exchanger with hot fluid cooling from $T_{h,i}$
to $T_{h,o}$ and cold fluid heating from $T_{c,i}$ to $T_{c,o}$, the
temperature differences at the two ends are $\Delta T_1 = T_{h,i} -
T_{c,o}$ and $\Delta T_2 = T_{h,o} - T_{c,i}$. The LMTD is
\[
\Delta T_{\text{LMTD}} = \frac{\Delta T_1 - \Delta T_2}{\ln(\Delta
T_1/\Delta T_2)},
\]
defined so that when both end differences are equal ($\Delta T_1 =
\Delta T_2$), the LMTD equals the common value (the limit of the
expression is $\Delta T_1$ as $\Delta T_2 \to \Delta T_1$). The heat-
transfer rate is then
\[
\dot{Q} = U A \Delta T_{\text{LMTD}},
\]
where $U$ is the overall heat-transfer coefficient (chapter~5) and
$A$ is the heat-transfer area.

The LMTD formula is derived from the differential energy balance
along the exchanger length, assuming constant $U$ and constant fluid
specific heats. The counter-flow case has the largest LMTD for given
inlet and outlet temperatures and is the thermodynamically preferred
configuration. The parallel-flow (co-flow) case has both fluids
entering at one end; the temperature difference falls toward the
exit, and the outlet hot temperature is necessarily greater than the
outlet cold temperature (the two fluids cannot cross). The LMTD
formula applies to parallel flow with $\Delta T_1 = T_{h,i} -
T_{c,i}$ and $\Delta T_2 = T_{h,o} - T_{c,o}$; the resulting LMTD is
smaller than the counter-flow value for the same external
temperatures.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-flow-temperature-profiles.tex}

For exchangers that are neither pure counter-flow nor pure parallel-
flow (shell-and-tube with multiple tube passes, cross-flow with both
streams unmixed, plate-and-frame), a \engterm{correction factor} $F
< 1$ is applied to the counter-flow LMTD:
\[
\dot{Q} = U A F \Delta T_{\text{LMTD,CF}}.
\]
Correction factors are tabulated as functions of two dimensionless
ratios: $P = (T_{c,o} - T_{c,i})/(T_{h,i} - T_{c,i})$ and $R =
(T_{h,i} - T_{h,o})/(T_{c,o} - T_{c,i})$ \cite{text:incropera2007}.
The correction factor must be $\geq 0.75$ for a practical design; a
lower value indicates the exchanger configuration cannot achieve the
specified terminal temperatures and either more tube passes (for
shell-and-tube) or a different exchanger type is needed.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-lmtd-correction-factor.tex}

\subsection{The log-mean from the differential balance}

The opening statement that the LMTD ``is derived from the differential energy
balance'' is worth carrying out, because the derivation shows why the mean is
logarithmic rather than arithmetic and exactly which assumptions it rests on. Take
an element of the exchanger that carries a slice of area $dA$. Let $\theta(A) = T_h
- T_c$ be the local difference between the two stream temperatures, written as a
function of the cumulative area $A$ measured from one end. The heat crossing the
slice is $d\dot{Q} = U\theta\,dA$, and this heat cools the hot stream and warms the
cold stream:
\[
dT_h = -\frac{d\dot{Q}}{C_h}, \qquad dT_c = \pm\frac{d\dot{Q}}{C_c},
\]
where $C_h = \dot{m}_h c_{p,h}$ and $C_c = \dot{m}_c c_{p,c}$ are the capacity
rates, and the sign on $dT_c$ is positive for counter-flow (the cold stream flows
opposite the area coordinate, so it cools as $A$ increases in the integration
sense used here) and negative for parallel-flow. The change in the local
difference along the slice is
\[
d\theta = dT_h - dT_c = -d\dot{Q}\left(\frac{1}{C_h} \pm \frac{1}{C_c}\right)
= -U\theta\,dA\left(\frac{1}{C_h} \pm \frac{1}{C_c}\right).
\]
Write $1/C^\ast = 1/C_h \pm 1/C_c$, with the plus sign for counter-flow and the
minus for parallel-flow. The slice equation separates,
\[
\frac{d\theta}{\theta} = -\frac{U}{C^\ast}\,dA,
\]
and because $U$, $C_h$, and $C_c$ are taken constant along the length, the right
side integrates to a straight line in $A$. Integrating from one end ($\theta =
\theta_1$, $A = 0$) to the other ($\theta = \theta_2$, $A = A_{\text{tot}}$),
\[
\ln\frac{\theta_2}{\theta_1} = -\frac{U A_{\text{tot}}}{C^\ast}.
\]
The driving difference therefore decays as a single exponential in the area
traversed, the curve drawn in figure \ref{fig:vol04ch06:deltat-decay}. To recover
the design equation, eliminate $C^\ast$. The overall energy balance gives
$\dot{Q} = C^\ast(\theta_1 - \theta_2)$ for both arrangements (the difference of
the two terminal driving differences, weighted by the combined capacity rate), and
the same $\dot{Q}$ must equal $U A_{\text{tot}}$ times whatever mean difference we
choose to define. Setting $\dot{Q} = U A_{\text{tot}}\,\Delta T_m$ and using the
integrated result to write $U A_{\text{tot}}/C^\ast = \ln(\theta_1/\theta_2)$,
\[
\dot{Q} = C^\ast(\theta_1 - \theta_2)
= \frac{U A_{\text{tot}}}{\ln(\theta_1/\theta_2)}(\theta_1 - \theta_2),
\]
so the mean that makes $\dot{Q} = U A_{\text{tot}}\,\Delta T_m$ exact is
\[
\Delta T_m = \frac{\theta_1 - \theta_2}{\ln(\theta_1/\theta_2)}
= \Delta T_{\text{LMTD}}.
\]
The log-mean is not a modelling choice but the consequence of integrating an
exponentially decaying driving difference: the heat transferred is the integral of
$U\theta\,dA$, and the area-average of a log-linearly varying $\theta$ is its
log-mean. The arithmetic mean $(\theta_1 + \theta_2)/2$ always exceeds the log-mean
(by the arithmetic-geometric inequality applied to the two end differences), so
sizing on the arithmetic mean undersizes the exchanger; the error is small when the
two end differences are within a factor of about two and grows without bound as one
of them approaches zero. Three assumptions carry the whole derivation: constant
$U$, constant specific heats, and no longitudinal conduction or heat loss to
ambient. Each is relaxed in the same way, by sectioning the exchanger into zones
within which the assumption holds and chaining the zones, which is what
rating software does for a condenser whose $U$ jumps at the dew point.

\subsection{Why the correction factor exists, and what sets its floor}

The exponential decay above holds cleanly only for pure counter-flow or pure
parallel-flow, where every fluid particle of one stream sees the same routing past
the other. A one-shell-pass, two-tube-pass unit violates this: half the tube
passes run counter to the shell flow and half run with it, so a given tube-side
particle alternates between favourable and unfavourable driving difference along
its path. A cross-flow unit violates it differently, because the two streams cross
at right angles and no single $\theta(A)$ describes the whole field. In both cases
the true mean driving difference is smaller than the counter-flow log-mean computed
from the same four terminal temperatures, and the ratio of the two is the
correction factor:
\[
F = \frac{\Delta T_m^{\text{actual}}}{\Delta T_{\text{LMTD,CF}}} \le 1.
\]
The factor is therefore a geometry penalty, the price the configuration pays for not
being pure counter-flow, and it is computed once for each arrangement by solving
that arrangement's two-dimensional temperature field, integrating to find the
actual mean, and dividing by the counter-flow log-mean. The result depends on the
four terminal temperatures only through the two dimensionless groups $P$ and $R$
defined above, which is why $F$ can be charted as $F(P, R)$ for each configuration
rather than recomputed for every duty. The group $P$ is a temperature
effectiveness, the fraction of the maximum possible rise the cold stream actually
achieves; $R$ is the ratio of the two capacity rates. The steep collapse of every
$F(P)$ curve near a configuration-dependent value of $P$ has a physical reading: it
is the point at which the requested cold-stream rise approaches what a single shell
pass can deliver, where the co-current portions of the path run out of driving
difference. Past that point no real $F$ exists, because the configuration cannot
reach the requested terminal temperatures at all. The working floor $F \ge 0.75$
keeps the design off the steep part of the curve, where a small terminal-temperature
error swings $F$ sharply, and the standard repair when a duty pushes $F$ below the
floor is to add a second shell in series, which restores more of the counter-flow
character to the path and lifts $F$ back toward unity. The chart of figure
\ref{fig:vol04ch06:lmtd-correction-factor} is the working form of $F(P,R)$ for the
commonest configuration.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-deltat-decay.tex}

A worked sizing fixes the method. Take a counter-flow exchanger that
must cool $3\,\text{kg/s}$ of water from $90\,\degC$ to $50\,\degC$
against $4\,\text{kg/s}$ of water entering at $20\,\degC$, with $U =
1200\,\si{\watt\per\meter\squared\per\kelvin}$. The duty is $\dot{Q}
= \dot{m}_h c_p \Delta T_h = 3 \times 4180 \times 40 = 501.6\,\text{kW}$.
The cold outlet follows from the cold-side balance: $T_{c,o} = 20 +
\dot{Q}/(\dot{m}_c c_p) = 20 + 501{,}600/(4 \times 4180) = 50.0\,\degC$.
The two end differences are $\Delta T_1 = 90 - 50 = 40\,\si{\kelvin}$
and $\Delta T_2 = 50 - 20 = 30\,\si{\kelvin}$, so $\Delta
T_{\text{LMTD}} = (40 - 30)/\ln(40/30) = 34.8\,\si{\kelvin}$. The area
is $A = \dot{Q}/(U \Delta T_{\text{LMTD}}) = 501{,}600/(1200 \times
34.8) = 12.0\,\text{m}^2$. The routine \texttt{lmtd\_sizing.py} carries
out this arithmetic and the parallel-flow comparison.

\section{NTU-effectiveness method}\pathtag{core}

The \engterm{number of transfer units}–\engterm{effectiveness} method
solves the inverse problem: given the heat-transfer area, the
overall coefficient, the two inlet temperatures, and the two mass-
flow rates, find the outlet temperatures and the heat-transfer rate.
The method is the working tool when the area is specified (a given
piece of equipment) and the operating temperatures are to be
predicted.

The two key dimensionless groups:
\begin{align*}
\text{NTU} &= \frac{U A}{C_{\min}}, \\
C_r &= \frac{C_{\min}}{C_{\max}},
\end{align*}
where $C_h = \dot{m}_h c_{p,h}$ and $C_c = \dot{m}_c c_{p,c}$ are the
two heat-capacity rates, and $C_{\min}$ and $C_{\max}$ are the smaller
and larger of the two. The effectiveness $\varepsilon$ is the actual
heat transfer divided by the maximum possible heat transfer
(approached if the exchanger had infinite area):
\[
\varepsilon = \frac{\dot{Q}}{\dot{Q}_{\max}} = \frac{\dot{Q}}{C_{\min}
(T_{h,i} - T_{c,i})}.
\]
The effectiveness is between $0$ (no heat transferred) and $1$ (one
stream reaches the other's inlet temperature). The relation between
$\varepsilon$, NTU, and $C_r$ depends on the exchanger configuration:

\begin{itemize}[leftmargin=*]
  \item Counter-flow: $\varepsilon = (1 - e^{-\text{NTU}(1 - C_r)})/
    (1 - C_r e^{-\text{NTU}(1 - C_r)})$ for $C_r \neq 1$; $\varepsilon
    = \text{NTU}/(1 + \text{NTU})$ for $C_r = 1$.
  \item Parallel-flow: $\varepsilon = (1 - e^{-\text{NTU}(1 + C_r)})/
    (1 + C_r)$.
  \item Cross-flow, both streams unmixed: $\varepsilon = 1 - \exp\{
    [\exp(-\text{NTU}^{0.78}\,C_r) - 1]\text{NTU}^{0.22}/C_r\}$
    (correlation, used widely in air-cooled and compact-exchanger
    design).
  \item Shell-and-tube, one shell pass, $2,4,\ldots$ tube passes:
    standard $\varepsilon$-NTU expression with intermediate complexity.
\end{itemize}

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-effectiveness-ntu.tex}

For an evaporator or condenser (where one fluid undergoes phase
change at constant temperature), $C_{\max} \to \infty$ on that side
and $C_r \to 0$, giving $\varepsilon = 1 - e^{-\text{NTU}}$ regardless
of the exchanger configuration. The simplification is one reason
phase-change exchangers are easier to design than sensible-heat
exchangers.

The NTU method's working power is that the relations are explicit
functions of NTU and $C_r$; given the area, the coefficient, and the
flows, the outlet temperatures follow by direct substitution. The
LMTD method requires an iteration when the area is given and the
temperatures are unknown.

\subsection{Deriving the parallel-flow relation from the differential balance}

The counter-flow effectiveness relation is derived in the exercises; the
parallel-flow case is shorter and shows the machinery in its simplest form. Start
again from the integrated balance of the differential derivation, now with the
parallel-flow sign, $1/C^\ast = 1/C_h + 1/C_c$. Taking the hot stream as
$C_{\min}$ so that $\text{NTU} = UA/C_{\min}$ and $C_r = C_{\min}/C_{\max}$, the
combined group is $UA/C^\ast = UA(1/C_h + 1/C_c) = \text{NTU}(1 + C_r)$. The
integrated balance reads
\[
\frac{\theta_2}{\theta_1} = e^{-\text{NTU}(1 + C_r)},
\]
where for parallel-flow both streams enter at the same end, so $\theta_1 = T_{h,i}
- T_{c,i}$ is the full inlet difference and $\theta_2 = T_{h,o} - T_{c,o}$ is the
outlet difference. Write the two outlet temperatures through the effectiveness:
$T_{h,o} = T_{h,i} - \varepsilon(T_{h,i} - T_{c,i})$ from the definition of
$\varepsilon$ on the $C_{\min}$ stream, and $T_{c,o} = T_{c,i} + \varepsilon
C_r(T_{h,i} - T_{c,i})$ from the cold-side balance. Subtract:
\[
\theta_2 = T_{h,o} - T_{c,o}
= (T_{h,i} - T_{c,i})\bigl[1 - \varepsilon - \varepsilon C_r\bigr]
= \theta_1\bigl[1 - \varepsilon(1 + C_r)\bigr].
\]
Equating the two expressions for $\theta_2/\theta_1$,
\[
1 - \varepsilon(1 + C_r) = e^{-\text{NTU}(1 + C_r)},
\]
and solving for the effectiveness,
\[
\varepsilon = \frac{1 - e^{-\text{NTU}(1 + C_r)}}{1 + C_r},
\]
the parallel-flow relation quoted above. Two limits read straight off it. As
$\text{NTU} \to \infty$ the exponential vanishes and $\varepsilon \to 1/(1 + C_r)$,
a ceiling strictly below one whenever $C_r > 0$: parallel-flow cannot deliver full
effectiveness no matter how much area is bought, because the two outlets must
approach the same temperature and that common temperature lies between the two
inlets. At $C_r = 1$ the ceiling is exactly $0.5$, the two streams meeting in the
middle. The counter-flow relation has no such ceiling below one, which is the
quantitative statement of why counter-flow is the thermodynamically preferred
arrangement: at every NTU it returns more effectiveness, and at large NTU it
approaches the full unity that parallel-flow can never reach. Figure
\ref{fig:vol04ch06:arrangement-effectiveness} sets the four common arrangements
side by side at a fixed $C_r$, and the gap between the top and bottom curves is the
duty an engineer buys or forfeits by the routing alone.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-arrangement-effectiveness.tex}

\subsection{The one-shell-pass relation and the cross-flow correlation}

The shell-and-tube arrangement with one shell pass and an even number of tube
passes has a closed-form effectiveness that is worth writing down, because it is
the relation behind the kerosene cooler's correction factor and the
one-shell-pass curve of the figure. The result, with $\gamma = \sqrt{1 + C_r^2}$,
is
\[
\varepsilon_1 = 2\left[(1 + C_r) + \gamma\,
\frac{1 + e^{-\text{NTU}\,\gamma}}{1 - e^{-\text{NTU}\,\gamma}}\right]^{-1},
\]
which collapses to the phase-change form $1 - e^{-\text{NTU}}$ as $C_r \to 0$ and
sits between the parallel-flow and counter-flow curves at finite $C_r$, as the
figure shows. The structure of the formula records its origin: the
$\sqrt{1 + C_r^2}$ is the effective decay rate of a path that runs half
counter-current and half co-current, the geometric blend of the $(1 - C_r)$ of
pure counter-flow and the $(1 + C_r)$ of pure parallel-flow. For $N$ shell passes
in series, each carrying the same $\varepsilon_1$, the overall effectiveness
follows from composing the passes through their shared $C_r$:
\[
\varepsilon = \frac{\bigl[(1 - \varepsilon_1 C_r)/(1 - \varepsilon_1)\bigr]^N - 1}
{\bigl[(1 - \varepsilon_1 C_r)/(1 - \varepsilon_1)\bigr]^N - C_r},
\]
which is the relation a designer uses to decide how many shells a hard duty needs.
The cross-flow arrangement with both streams unmixed has no elementary closed form
at all. The exact effectiveness is a double integral of the coupled
two-dimensional field whose series solution is a sum of modified Bessel functions,
and design practice replaces it with the compact correlation
\[
\varepsilon = 1 - \exp\!\left\{\frac{\text{NTU}^{0.22}}{C_r}
\left[e^{-C_r\,\text{NTU}^{0.78}} - 1\right]\right\},
\]
accurate to about one percent across the engineering range. The lesson the four
relations carry together is that effectiveness is geometry-specific: every routing
has its own $\varepsilon(\text{NTU}, C_r)$, and the single most common error in
rating an existing unit is to apply the counter-flow relation to a cross-flow or
shell-and-tube exchanger, which over-states the effectiveness and the duty.

\subsection{The equivalence of the two methods}

The principle box below states that LMTD and NTU are duals; the equivalence is an
identity, not an analogy, and seeing it removes any worry that the two routes might
give different answers. Begin from the NTU side with counter-flow. The number of
transfer units is $\text{NTU} = UA/C_{\min}$, and the integrated balance gave
$\ln(\theta_1/\theta_2) = UA(1/C_{\min} - 1/C_{\max}) = \text{NTU}(1 - C_r)$. The
duty written through the effectiveness is $\dot{Q} = \varepsilon C_{\min}\,\Delta
T_{\max}$ with $\Delta T_{\max} = T_{h,i} - T_{c,i}$, while the duty written
through the log-mean is $\dot{Q} = UA\,\Delta T_{\text{LMTD}}$. Dividing the two,
\[
\frac{\Delta T_{\text{LMTD}}}{\Delta T_{\max}}
= \frac{\varepsilon C_{\min}}{UA}
= \frac{\varepsilon}{\text{NTU}}.
\]
Now express $\Delta T_{\text{LMTD}}$ through the two end differences. For
counter-flow $\theta_1 = T_{h,i} - T_{c,o} = \Delta T_{\max}(1 - \varepsilon C_r)$
and $\theta_2 = T_{h,o} - T_{c,i} = \Delta T_{\max}(1 - \varepsilon)$, so
\[
\Delta T_{\text{LMTD}}
= \frac{\theta_1 - \theta_2}{\ln(\theta_1/\theta_2)}
= \Delta T_{\max}\,\frac{\varepsilon(1 - C_r)}{\ln\!\dfrac{1 - \varepsilon C_r}{1 - \varepsilon}}.
\]
Substituting into the ratio and cancelling $\Delta T_{\max}$,
\[
\frac{\varepsilon}{\text{NTU}}
= \frac{\varepsilon(1 - C_r)}{\ln\!\dfrac{1 - \varepsilon C_r}{1 - \varepsilon}},
\quad\text{so}\quad
\text{NTU}(1 - C_r) = \ln\frac{1 - \varepsilon C_r}{1 - \varepsilon},
\]
which rearranges to exactly the counter-flow effectiveness relation derived in the
exercises. The two methods are the same equation solved for different unknowns: fix
the temperatures and solve for the area (LMTD), or fix the area and solve for the
temperatures (NTU). No information is added or lost in passing between them, and a
calculation that uses both must close on itself, which is the consistency check the
worked-examples section runs.

\subsection{Networks of streams: the multi-stream balance}

A single exchanger handles two streams; a process train handles many, and the
balance that governs the train is the same first law applied to a control volume
that contains several exchangers. For a network of $n$ hot streams and $m$ cold
streams exchanging heat among themselves and with external utilities, steady
operation requires that the total enthalpy entering equals the total enthalpy
leaving. Writing the duty of exchanger $k$ as $\dot{Q}_k$ and the external hot and
cold utility as $\dot{Q}_{\text{hot}}$ and $\dot{Q}_{\text{cold}}$, the overall
balance is
\[
\sum_{i=1}^{n} C_{h,i}(T_{h,i}^{\text{in}} - T_{h,i}^{\text{out}})
+ \dot{Q}_{\text{hot}}
= \sum_{j=1}^{m} C_{c,j}(T_{c,j}^{\text{out}} - T_{c,j}^{\text{in}})
+ \dot{Q}_{\text{cold}},
\]
the heat given up by the hot streams and the hot utility equalling the heat taken
up by the cold streams and the cold utility. Each internal exchanger contributes a
$\dot{Q}_k$ that appears once as a loss on its hot stream and once as a gain on its
cold stream, so the internal duties cancel from the overall balance and only the
boundary terms remain. That cancellation is what lets pinch analysis target the
minimum utilities before any exchanger is placed: the composite-curve construction
of section 6.6 is this balance read interval by interval in temperature. The local
counterpart is the rule that at any internal junction where streams mix or split,
the mixed enthalpy is the flow-weighted sum of the branch enthalpies, the same rule
the bypass worked example uses to find a remixed outlet. A network solver assembles
one $\varepsilon$-NTU or LMTD relation per exchanger and one balance per stream
junction, and solves the coupled set; the two-stream relations of this chapter are
the building block, and the multi-stream balance is the frame that holds them
together.

\begin{principle}[LMTD and NTU-effectiveness are duals]
The LMTD method takes terminal temperatures and gives area; the NTU
method takes area and gives terminal temperatures. Both are
expressions of the same underlying energy balance on the exchanger,
parameterised differently. The choice between them is set by which
quantities are specified in the design problem: LMTD for sizing a
new exchanger to meet a specified duty, NTU-effectiveness for
rating an existing exchanger under new operating conditions.
\end{principle}

\section{Extended surfaces and the finned-surface coefficient}\pathtag{core}

A gas side runs a low film coefficient, often one to two orders of magnitude
below the liquid side it works against, so the gas-side resistance dominates the
series network and the cheapest way to relieve it is to add area where it is
needed. Fins do exactly that: they multiply the gas-side area at the cost of a
little extra metal, and the accounting that keeps the gain honest is the algebra
of fin efficiency and overall surface efficiency. The earlier worked example
quoted the result; we now derive it from the fin conduction equation, because the
third case study rests on it and because the same algebra recurs wherever
extended surface appears.

\subsection{The single-fin efficiency}

Take a straight fin of uniform cross-section: thickness $t$, width $w$ (into the
page), length $L$ from root to tip, thermal conductivity $k$, exposed to a fluid
at $T_\infty$ through a film coefficient $h$. Let $\theta(x) = T(x) - T_\infty$ be
the local excess temperature. A slice of the fin between $x$ and $x + dx$
conducts heat in at $x$, conducts it out at $x + dx$, and sheds the difference to
the fluid through its perimeter $P = 2(w + t) \approx 2w$ for a thin fin. The
steady energy balance on the slice is
\[
k A_c \frac{d^2\theta}{dx^2} = h P\,\theta,
\]
with $A_c = wt$ the conduction cross-section. Writing $m^2 = hP/(kA_c)$ this is
$\theta'' = m^2\theta$, whose solution with the root condition $\theta(0) =
\theta_b$ and an adiabatic tip ($\theta'(L) = 0$, the standard idealisation for a
fin whose tip loss is folded into a corrected length) is
\[
\frac{\theta(x)}{\theta_b} = \frac{\cosh[m(L - x)]}{\cosh(mL)}.
\]
The temperature falls from the root toward the tip along a hyperbolic cosine,
drawn in figure \ref{fig:vol04ch06:fin-temperature-decay}. The heat the fin
actually sheds equals the heat conducted into its root,
\[
\dot{Q}_{\text{fin}} = -kA_c \left.\frac{d\theta}{dx}\right|_{x=0}
= kA_c\, m\,\theta_b \tanh(mL) = \sqrt{hPkA_c}\;\theta_b \tanh(mL).
\]
An ideal fin, held everywhere at the root temperature, would shed
$\dot{Q}_{\text{ideal}} = h(PL)\theta_b$ over its whole surface area $PL$. The
\engterm{fin efficiency} is the ratio,
\[
\eta_f = \frac{\dot{Q}_{\text{fin}}}{\dot{Q}_{\text{ideal}}}
= \frac{\sqrt{hPkA_c}\,\tanh(mL)}{hPL}
= \frac{\tanh(mL)}{mL},
\]
since $mL = L\sqrt{hP/(kA_c)}$ cancels the prefactor. The efficiency is one when
$mL \to 0$ (a short, fat, highly conductive fin behaves like the ideal isothermal
fin) and falls toward $1/(mL)$ for large $mL$ (a long, thin, poorly conducting
fin barely heats its tip). The group $mL$ is the single dimensionless number that
governs the fin, and the design lives in the range where $\eta_f$ sits between
about $0.7$ and $0.95$: lower than that wastes metal on a fin whose tip does no
work, higher than that means the fin is too short to be buying much area.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-fin-temperature-decay.tex}

\subsection{From single fin to overall surface efficiency}

A finned surface is part bare tube wall (the prime surface between the fins) and
part fin. The bare part runs at the root temperature and is fully effective; the
fin part runs at $\eta_f$ of the root driving difference. Let $A_f$ be the total
fin area and $A$ the total air-side area (bare plus fin), so the bare area is
$A - A_f$. The total heat shed by the finned surface at root excess $\theta_b$ is
\[
\dot{Q} = h(A - A_f)\theta_b + h A_f \eta_f \theta_b
= h A\,\theta_b\left[1 - \frac{A_f}{A}(1 - \eta_f)\right].
\]
The bracket is the \engterm{overall surface efficiency},
\[
\eta_o = 1 - \frac{A_f}{A}(1 - \eta_f),
\]
the factor that multiplies the bare film coefficient when the air-side
conductance is referred to the total finned area: the effective conductance is
$\eta_o h A$ rather than $hA$. The overall surface efficiency is a weighted blend
of the fully-effective bare surface and the partly-effective fin surface, and it
is always between $\eta_f$ and one. A design that treats the entire finned area as
if it ran at the root temperature overcounts the air-side conductance by the
factor $1/\eta_o$, which for a typical fin field ($A_f/A \approx 0.9$,
$\eta_f \approx 0.85$) is a silent fifteen-percent error in the dominant
resistance.

\subsection{The overall coefficient on the finned area}

The air-side conductance now joins the wall and the tube-side film in series, and
because the two sides have very different areas the network must be referred to
one of them. Referring everything to the total air-side (finned) area $A_o$, the
overall coefficient $U_o$ satisfies
\[
\frac{1}{U_o} = \frac{1}{\eta_o h_o}
+ R_{f,o}
+ \frac{A_o \ln(d_o/d_i)}{2\pi k_w L_t}
+ \frac{A_o}{A_i} R_{f,i}
+ \frac{A_o}{A_i}\frac{1}{h_i},
\]
where $h_o$ is the air-side film coefficient on the finned surface, $h_i$ the
tube-side (process) film, $A_i = \pi d_i L_t$ the inside bare area, and
$A_o/A_i$ the area ratio that grows large when the finning is heavy. The two
features that distinguish this from the bare-tube network of chapter~5 are the
$\eta_o$ multiplying the air-side film and the large $A_o/A_i$ amplifying every
tube-side term. A heavy fin field with $A_o/A_i$ of fifteen to twenty-five turns a
respectable tube-side film into a small contributor referred to the air side,
which is the quantitative reason the air-side film, even multiplied by all that
area, still tends to govern an air cooler. The algebra is the tool the third case
study uses to build the coefficient term by term, the same discipline the
shell-and-tube case applied to a bare-tube network.

\begin{mastery}
The adiabatic-tip assumption behind $\eta_f = \tanh(mL)/mL$ is an idealisation;
a real fin sheds a little heat from its tip. The standard repair is the corrected
length $L_c = L + t/2$, which adds a strip of area equal to the tip and recovers
the tip loss to within a fraction of a percent for the thin fins of practice,
so design uses $\eta_f = \tanh(mL_c)/mL_c$ without solving the convective-tip
boundary condition. A second subtlety is that fins are not always straight: annular
(radial) fins on a round tube, the common geometry on air-cooler tubing, obey the
same conduction equation in cylindrical coordinates, whose solution is a
combination of modified Bessel functions $I_0, I_1, K_0, K_1$ rather than a
hyperbolic cosine. The efficiency of an annular fin is read from a chart or a
Bessel-function expression as a function of the radius ratio and the same $mL$
group, and it sits a little below the straight-fin value for the same height
because the outer annulus has more area to feed at the cooler end. Design practice
charts annular-fin efficiency once and reuses it; the straight-fin formula is the
mental model and the conservative first estimate.
\end{mastery}

\section{Worked examples}\pathtag{standard}

Two worked rating problems show the NTU method end to end and expose
the points where a calculation goes wrong.

\textbf{Rating an air-cooled oil exchanger.} An existing fin-fan cooler
has $A = 200\,\text{m}^2$ and $U = 50\,\si{\watt\per\meter\squared\per
\kelvin}$. It cools $5\,\text{kg/s}$ of oil ($c_p = 2.1\,\text{kJ}/
(\text{kg}\cdot\si{\kelvin})$) entering at $140\,\degC$ with $20\,
\text{kg/s}$ of air ($c_p = 1.005\,\text{kJ}/(\text{kg}\cdot\si{
\kelvin})$) entering at $30\,\degC$. The capacity rates are $C_{\text{oil}}
= 5 \times 2100 = 10{,}500\,\si{\watt\per\kelvin}$ and $C_{\text{air}}
= 20 \times 1005 = 20{,}100\,\si{\watt\per\kelvin}$, so $C_{\min} =
10{,}500$ (the oil), $C_r = 10{,}500/20{,}100 = 0.52$, and $\text{NTU}
= UA/C_{\min} = 50 \times 200/10{,}500 = 0.95$. For a cross-flow
exchanger with both streams unmixed the correlation gives $\varepsilon
\approx 0.53$. The duty is $\dot{Q} = \varepsilon C_{\min}(T_{h,i} -
T_{c,i}) = 0.53 \times 10{,}500 \times 110 = 610\,\text{kW}$. The oil
leaves at $140 - 610{,}000/10{,}500 = 82\,\degC$ and the air at $30 +
610{,}000/20{,}100 = 60\,\degC$. The trap here is the choice of
$C_{\min}$: the smaller capacity rate sets both the NTU denominator and
the maximum possible duty, and using the oil's value (not the air's) is
what makes the answer correct. The routine
\texttt{ntu\_effectiveness.py} reproduces these numbers.

\textbf{When LMTD and NTU disagree on effort.} For the same exchanger,
suppose instead that the four temperatures were specified and the area
were the unknown. The LMTD route would need the outlet temperatures
first, which here are exactly what the NTU route produces, so solving
for area from specified temperatures would require iterating the
unknown outlet, computing the LMTD, and back-solving the area until the
energy balance closed. The NTU route gives the outlets in one pass.
The lesson is to read the problem before choosing the method: area
unknown and temperatures known favours LMTD; area known and
temperatures unknown favours NTU.

\textbf{Fouling-driven decline of an existing exchanger.} A cooling-water-
to-process exchanger left service clean at $U_{\text{clean}} = 1100\,
\si{\watt\per\meter\squared\per\kelvin}$, area $A = 18\,\text{m}^2$,
cooling $4\,\text{kg/s}$ of process fluid ($c_p = 2.4\,\text{kJ}/(\text{
kg}\cdot\si{\kelvin})$) from $95\,\degC$ against $6\,\text{kg/s}$ of
water entering at $28\,\degC$. We want the duty when clean and the duty
after the water-side fouling resistance has grown to $R_f = 0.0006\,
\text{m}^2\cdot\si{\kelvin}/\text{W}$. The capacity rates are $C_p = 4
\times 2400 = 9600\,\si{\watt\per\kelvin}$ and $C_w = 6 \times 4180 =
25{,}080\,\si{\watt\per\kelvin}$, so $C_{\min} = 9600$ (the process
fluid) and $C_r = 9600/25{,}080 = 0.38$. Clean, the number of transfer
units is $\text{NTU} = U_{\text{clean}} A/C_{\min} = 1100 \times 18/
9600 = 2.06$; the counter-flow relation gives $\varepsilon = 0.85$ and a
duty of $\dot{Q} = \varepsilon C_{\min}(T_{h,i} - T_{c,i}) = 0.85 \times
9600 \times 67 = 547\,\text{kW}$, leaving the process fluid at $95 -
547{,}000/9600 = 38\,\degC$. Fouling adds resistance in series: $1/
U_{\text{dirty}} = 1/1100 + 0.0006 = 0.000909 + 0.0006 = 0.001509$, so
$U_{\text{dirty}} = 663\,\si{\watt\per\meter\squared\per\kelvin}$. The
new NTU is $663 \times 18/9600 = 1.24$, the effectiveness falls to
$\varepsilon = 0.69$, and the duty drops to $0.69 \times 9600 \times 67
= 444\,\text{kW}$, with the process fluid now leaving at $49\,\degC$
rather than $38\,\degC$. A fouling layer that adds two-thirds of the
clean wall-plus-film resistance has cut the duty by $19\,\%$ and pushed
the process outlet up by $11\,\si{\kelvin}$. The lesson the NTU route
makes plain: effectiveness is concave in NTU, so an exchanger sized for
a high clean effectiveness (here $0.85$) sheds duty less steeply under
fouling than one run near the linear part of the curve, which is one
quantitative argument for the over-design margin of the failure
section. The routine \texttt{ntu\_effectiveness.py} returns both
operating points.

\textbf{An economiser recovering boiler flue-gas heat.} A boiler
exhausts $8\,\text{kg/s}$ of flue gas ($c_p = 1.10\,\text{kJ}/(\text{kg}
\cdot\si{\kelvin})$) at $260\,\degC$. An economiser preheats $5\,\text{
kg/s}$ of feedwater ($c_p = 4.20\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$)
from $40\,\degC$ before it enters the boiler, recovering heat that would
otherwise go up the stack. The economiser is a finned cross-flow unit
with $UA = 14{,}000\,\text{W}/\si{\kelvin}$. The capacity rates are
$C_g = 8 \times 1100 = 8800\,\si{\watt\per\kelvin}$ and $C_w = 5 \times
4200 = 21{,}000\,\si{\watt\per\kelvin}$, so $C_{\min} = 8800$ (the gas),
$C_r = 8800/21{,}000 = 0.42$, and $\text{NTU} = 14{,}000/8800 = 1.59$.
The cross-flow-both-unmixed correlation gives $\varepsilon \approx 0.71$.
The recovered duty is $\dot{Q} = 0.71 \times 8800 \times (260 - 40) =
1.37\,\text{MW}$. The gas leaves at $260 - 1{,}370{,}000/8800 = 104\,
\degC$ and the feedwater enters the boiler at $40 + 1{,}370{,}000/
21{,}000 = 105\,\degC$. The recovered $1.37\,\text{MW}$ is heat the
boiler no longer has to supply from fuel, a direct efficiency gain. The
binding constraint on pushing the gas outlet lower is the acid dew point:
sulphur in the fuel forms sulphuric acid that condenses on the cold end
below roughly $120$ to $150\,\degC$ and corrodes the economiser, so the
$104\,\degC$ gas outlet here would demand a corrosion-resistant cold-end
material or a higher feedwater inlet temperature to lift the metal
temperature out of the condensing band. The economiser shows the
recurring tension of heat recovery: the last increment of recovered heat
is the cheapest in fuel and the most expensive in corrosion risk.

\textbf{A finned-tube air cooler and the surface-efficiency factor.} Extended
surface buys area cheaply on a gas side, where the film coefficient is low, but
the fins do not work as hard as the bare tube wall because heat must conduct out
along the fin before it leaves to the air. The accounting is the
\engterm{overall surface efficiency} $\eta_o$, which multiplies the air-side
film coefficient when it is referred to the total (finned plus bare) area. Take
an air cooler with $5\,\text{kg/s}$ of process gas to be cooled, an air-side
film coefficient $h_o = 60\,\si{\watt\per\meter\squared\per\kelvin}$ on a finned
surface whose fin area is $85\,\%$ of the total air-side area, and a single-fin
efficiency $\eta_f = 0.82$ for the chosen fin height and metal. The surface
efficiency is $\eta_o = 1 - (A_f/A)(1 - \eta_f) = 1 - 0.85\times(1 - 0.82) =
0.847$, so the effective air-side conductance is $\eta_o h_o = 0.847\times 60 =
50.8\,\si{\watt\per\meter\squared\per\kelvin}$ rather than the bare $60$. The
$15\,\%$ shortfall is the price of moving heat along the fin, and it is the
number a designer who treats the whole finned area as if it ran at the root
temperature would silently lose. The tube-side process film, referred to the
same total air-side area through the area ratio $A_o/A_i$, then adds in series
with $1/(\eta_o h_o)$ to give the overall coefficient on the air-side basis. Fin
efficiency falls as the fins are made taller or thinner or out of a poorer
conductor, so the design trades fin area (cheap) against fin efficiency (which
erodes the value of that area), and the surface-efficiency factor is what keeps
the trade honest.

\textbf{Two exchangers in series and the additivity of transfer units.} A duty
that one shell cannot reach is often split between two shells in series, and the
NTU method composes them cleanly. Suppose a process stream with $C_{\min} =
6000\,\si{\watt\per\kelvin}$ passes through two counter-flow shells, the first
with $U_1 A_1 = 9000\,\text{W}/\si{\kelvin}$ and the second with $U_2 A_2 =
6000\,\text{W}/\si{\kelvin}$, each against a cooling stream large enough that
$C_r \approx 0$ for the pair (a near-isothermal coolant). With $C_r = 0$ the
counter-flow effectiveness is $\varepsilon = 1 - e^{-\text{NTU}}$, and transfer
units add along the path because the exponentials multiply: the first shell has
$\text{NTU}_1 = 9000/6000 = 1.5$ and the second $\text{NTU}_2 = 6000/6000 =
1.0$, so the series total is $\text{NTU} = 2.5$ and the overall effectiveness is
$\varepsilon = 1 - e^{-2.5} = 0.918$. Treating the two shells as one exchanger
of $\text{NTU} = 2.5$ gives the same answer as marching the outlet of the first
into the inlet of the second, which is the check worth doing: $\varepsilon_1 = 1
- e^{-1.5} = 0.777$ leaves the stream at a fraction $1 - 0.777 = 0.223$ of its
initial approach, the second shell leaves $1 - (1 - e^{-1.0}) = e^{-1.0} =
0.368$ of what reaches it, and the product $0.223\times 0.368 = 0.082$ is the
surviving approach, so the combined effectiveness is $1 - 0.082 = 0.918$,
matching the additive-NTU result. The additivity holds only at $C_r = 0$; for
finite $C_r$ the shells must be composed through their outlet temperatures, one
into the next, because the capacity-rate ratio changes the shape of each
effectiveness relation.

\textbf{An LMTD-versus-NTU consistency check.} The two methods are duals, so a
calculation that uses both must close on itself, and forcing the closure catches
arithmetic slips. Take the counter-flow water-to-water sizing of the LMTD
section: $\dot{Q} = 501.6\,\text{kW}$, $U = 1200\,\si{\watt\per\meter\squared
\per\kelvin}$, area $A = 12.0\,\text{m}^2$, hot side $C_h = 3\times 4180 =
12{,}540\,\si{\watt\per\kelvin}$, cold side $C_c = 4\times 4180 = 16{,}720\,
\si{\watt\per\kelvin}$. The NTU side reads $C_{\min} = 12{,}540$, $C_r = 12{,}540
/16{,}720 = 0.75$, and $\text{NTU} = UA/C_{\min} = 1200\times 12.0/12{,}540 =
1.148$. The counter-flow relation gives $\varepsilon = (1 - e^{-1.148(1 -
0.75)})/(1 - 0.75\,e^{-1.148(1 - 0.75)})$; the exponent is $-0.287$, so $e^{-
0.287} = 0.751$ and $\varepsilon = (1 - 0.751)/(1 - 0.75\times 0.751) = 0.249/
0.437 = 0.570$. The predicted duty is $\dot{Q} = \varepsilon C_{\min}(T_{h,i} -
T_{c,i}) = 0.570\times 12{,}540\times (90 - 20) = 500\,\text{kW}$, within
rounding of the $501.6\,\text{kW}$ the LMTD route started from. The two methods
agree because they are the same energy balance read in two directions; a
disagreement of more than a percent or two means a number was dropped, most
often the wrong stream taken as $C_{\min}$ or a correction factor applied on one
side but not the other.

\begin{mastery}
The cross-flow effectiveness correlation
$\varepsilon = 1 - \exp\{[\exp(-\text{NTU}^{0.78} C_r) - 1]
\text{NTU}^{0.22}/C_r\}$ is not an exact solution. The exact treatment
of cross-flow with both streams unmixed is a two-dimensional partial
differential equation whose solution is an infinite series of modified
Bessel functions, worked out for the single-pass case in the
mid-twentieth century. The compact closed form above reproduces the
exact series to within about one percent over the engineering range of
NTU and $C_r$, which is why design practice uses it and reserves the
series for verification. The deeper point is that effectiveness
relations are geometry-specific: every flow arrangement has its own
$\varepsilon(\text{NTU}, C_r)$, and substituting the counter-flow
relation for a cross-flow unit is a common and quietly large error.
\end{mastery}

\textbf{Rating an induced-draft cooling tower by range and approach.} A
direct-contact tower does not have an LMTD or a single $U$ because the heat and
mass transfer happen together at the air-water interface, so a tower is rated on
three numbers instead: the range, the approach, and the Merkel transfer-unit
demand it sets. Take a tower that must cool $80\,\text{kg/s}$ of circulating
water from a hot return of $38\,\degC$ to a cold supply of $29\,\degC$ against
ambient air at a design wet-bulb of $24\,\degC$. The range is the water
temperature drop, $\text{range} = 38 - 29 = 9\,\si{\kelvin}$, the duty the tower
performs; the approach is the gap the cooled water still leaves above the
wet-bulb, $\text{approach} = 29 - 24 = 5\,\si{\kelvin}$, the measure of how hard
the tower is working. The wet-bulb temperature is the thermodynamic floor: the
water cannot be cooled below it by any amount of air or fill, because the
wet-bulb is the temperature at which evaporation and sensible heat exchange with
the air come into balance. Figure
\ref{fig:vol04ch06:cooling-tower-range-approach} draws the water cooling down the
fill toward that floor with the range and the approach marked. The heat the
tower rejects is $\dot{Q} = \dot{m}_w c_p\,\text{range} = 80\times 4180\times 9 =
3.01\,\text{MW}$, almost all of it carried away as the latent heat of the small
fraction of water that evaporates: with the latent heat of water near $2.43\,
\text{MJ/kg}$, the evaporative loss is roughly $\dot{m}_{\text{evap}} \approx
\dot{Q}/h_{fg} = 3.01\times 10^6/2.43\times 10^6 = 1.24\,\text{kg/s}$, about
$1.5\,\%$ of the circulating flow, the standard rule that a tower evaporates
about $1\,\%$ of its flow per $7\,\si{\kelvin}$ of range. The transfer-unit
demand follows the Merkel integral, $\text{NTU}_{\text{Merkel}} = \int_{T_{\text{
cold}}}^{T_{\text{hot}}} c_p\,dT/(h_{s} - h_{a})$, the water-temperature integral
of the reciprocal enthalpy difference between saturated air at the local water
temperature and the bulk air; for this range and approach against a $24\,\degC$
wet-bulb the integral evaluates to roughly $\text{NTU}_{\text{Merkel}} \approx
1.4$, which the tower vendor matches with fill height and the air-to-water flow
ratio. The makeup water replaces three losses: evaporation ($1.24\,\text{kg/s}$),
drift (a small fraction carried out as droplets, held below $0.005\,\%$ by
modern eliminators), and blowdown (purged to keep dissolved solids from
concentrating as pure water evaporates away). The approach is what binds the
size: halving it from $5$ to $2.5\,\si{\kelvin}$ would push the Merkel integral
up sharply because the enthalpy difference near the cold end collapses toward
zero, the same area-hungry behaviour the steam condenser showed near its cold-end
approach, now read on a direct-contact unit.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-cooling-tower-range-approach.tex}

\begin{mastery}
A cooling tower is a simultaneous heat and mass exchanger, and treating it as if
it transferred sensible heat alone underestimates its capacity by a wide margin.
The driving potential is not the temperature difference between water and air but
the enthalpy difference between saturated air at the water-surface temperature
and the bulk air, which is the form Merkel's analysis takes. Two consequences
follow. First, the tower's performance depends on the air wet-bulb, not its
dry-bulb, because the wet-bulb fixes the bulk-air enthalpy that the saturated
interface air is driving against; a hot dry climate cools better than a cool
humid one at the same dry-bulb. Second, the same tower delivers different cold
water on different days as the wet-bulb moves, so a process that needs a
guaranteed supply temperature must be designed against the worst-case (highest)
wet-bulb of its location, which for many sites is a number exceeded only a few
hours a year. The Merkel model itself rests on simplifying assumptions, a unit
Lewis number and a neglected evaporative loss from the water flow, that the more
elaborate Poppe formulation removes; design practice rates on Merkel and reserves
Poppe for cases where the evaporation correction matters, the same closed-form
versus exact split the cross-flow effectiveness correlation showed.
\end{mastery}

\textbf{Sizing a single fin and reading its efficiency.} The fin algebra is most
concrete when a single fin is sized. Take an aluminium plate fin
($k = 200\,\si{\watt\per\meter\per\kelvin}$) of thickness $t = 0.4\,\si{\milli
\meter}$ and height $L = 16\,\si{\milli\meter}$, in an air stream with film
coefficient $h = 60\,\si{\watt\per\meter\squared\per\kelvin}$. For a thin straight
fin the perimeter-to-cross-section group reduces to $m = \sqrt{hP/(kA_c)} \approx
\sqrt{2h/(kt)}$, since $P/A_c = 2w/(wt) = 2/t$. Evaluating, $m = \sqrt{2\times
60/(200\times 0.0004)} = \sqrt{120/0.08} = \sqrt{1500} = 38.7\,\text{m}^{-1}$.
Using the corrected length $L_c = L + t/2 = 16.2\,\si{\milli\meter}$, the fin
group is $mL_c = 38.7\times 0.0162 = 0.627$, so the efficiency is $\eta_f =
\tanh(0.627)/0.627 = 0.556/0.627 = 0.887$. The fin runs at $89\,\%$ of an ideal
isothermal fin. Repeat with the same fin in steel
($k = 50\,\si{\watt\per\meter\per\kelvin}$): $m$ rises by $\sqrt{200/50} = 2$ to
$77.4\,\text{m}^{-1}$, $mL_c = 1.25$, and $\eta_f = \tanh(1.25)/1.25 =
0.848/1.25 = 0.678$. The poorer conductor drops the efficiency from $0.89$ to
$0.68$, which is the quantitative reason air-cooler fins are aluminium and not
steel: the cheaper metal would shed a fifth less heat per unit fin area. The
lesson is that fin efficiency is set by the group $mL$, and the conductivity sits
inside it under a square root, so halving $k$ does not halve the efficiency, it
moves $mL$ by $\sqrt{2}$ and the efficiency by the slope of $\tanh(mL)/mL$ at the
operating point.

\textbf{The optimum cooling-water rise and the area it buys.} A condenser or
cooler against cooling water carries a hidden optimisation in the choice of water
temperature rise. A small rise means a large water flow, a low pumping-feasible
velocity, more cooling-water pumping, and a large LMTD that shrinks the area; a
large rise means a small water flow, a small LMTD that swells the area, and a
water outlet that may cross the scaling limit. Take the kerosene cooler duty
($\dot{Q} = 2.20\,\text{MW}$, kerosene $160 \to 60\,\degC$, water in at
$30\,\degC$) and read the area against the chosen water rise. At a $10\,\si{\kelvin}$
rise the water leaves at $40\,\degC$, the terminal differences are $\Delta T_1 =
160 - 40 = 120$ and $\Delta T_2 = 60 - 30 = 30\,\si{\kelvin}$, the LMTD is
$(120 - 30)/\ln(120/30) = 90/1.386 = 64.9\,\si{\kelvin}$, and at $U = 407$ the
area is $2.20\times 10^6/(407\times 64.9) = 83\,\text{m}^2$ before the correction
factor. At a $22\,\si{\kelvin}$ rise (the case study's $52\,\degC$ outlet) the
LMTD falls to $61.0\,\si{\kelvin}$ and the bare area rises to $89\,\text{m}^2$. At
a $30\,\si{\kelvin}$ rise the water would leave at $60\,\degC$, equal to the
kerosene outlet, driving $\Delta T_2 \to 0$ and the area to infinity, the same
pinch the parallel-flow exercise showed. The water flow meanwhile scales as the
inverse of the rise: $\dot{Q}/(c_p\Delta T) = 2.20\times 10^6/(4180\times \Delta
T)$ gives $53\,\text{kg/s}$ at $10\,\si{\kelvin}$, $24\,\text{kg/s}$ at
$22\,\si{\kelvin}$, and $17.5\,\text{kg/s}$ at $30\,\si{\kelvin}$. The area is
flat across the practical range while the water flow, and with it the pumping and
the cooling-tower load, falls steeply, so the design pushes the rise as high as
the scaling limit allows, which is exactly why the case study sat the water outlet
at its $52\,\degC$ ceiling. The optimisation is read off the two opposing slopes:
area gently rising, water flow steeply falling, with the scaling limit the wall
that stops the rise.

\textbf{A bypass that defeats a good exchanger.} Field performance often falls
short of the rating not because the exchanger is undersized but because flow
bypasses it. Suppose the kerosene cooler is rated to leave kerosene at
$60\,\degC$, but a control bypass routes a fraction $f$ of the hot kerosene around
the shell and remixes it with the cooled stream. With $f = 0.15$ bypassed, the
$85\,\%$ that passes the shell is cooled close to design, say to $58\,\degC$,
while the bypassed $15\,\%$ stays at $160\,\degC$. The remixed outlet is the
flow-weighted average, $T_{\text{mix}} = 0.85\times 58 + 0.15\times 160 =
49.3 + 24.0 = 73.3\,\degC$, far above the $60\,\degC$ target despite the shell
doing its job on the flow it sees. The diagnostic signature distinguishes this
from fouling: a fouled exchanger shows a high outlet with a normal or slightly
raised pressure drop and a depressed coefficient when the duty is back-calculated
on the full flow, whereas a bypass shows a high outlet with a normal coefficient
on the flow that actually passes the shell and a tube-side pressure drop lower
than design because less flow is going through. The fix is to find and close the
bypass, not to clean or enlarge the exchanger, and the lesson is to confirm that
the flow the rating assumes is the flow the exchanger actually receives before
concluding the unit is at fault.

\textbf{Choosing the tube count from a velocity target rather than from area
alone.} The kerosene case rounded the tube count from the area and then found the
tube-side velocity uncomfortably low. The cleaner route runs the other way: fix a
target velocity, let it set the tubes per pass, and check that the resulting area
meets the duty. Take the same kerosene flow ($10\,\text{kg/s}$, $\rho =
800\,\text{kg/m}^3$) and aim for a tube-side velocity of $1.2\,\text{m/s}$, in the
middle of the liquid band. The required flow area per pass is $\dot{m}/(\rho u) =
10/(800\times 1.2) = 0.0104\,\text{m}^2$. With $d_i = 15\,\si{\milli\meter}$ tubes
of cross-section $(\pi/4)(0.015)^2 = 1.77\times 10^{-4}\,\text{m}^2$, the tubes per
pass are $0.0104/1.77\times 10^{-4} = 59$, so a two-pass unit carries $118$ tubes,
far fewer than the $372$ the area-first route gave at the lower velocity. Fewer
tubes means each must be longer to make the area: at $A_d = 108\,\text{m}^2$ and
$0.0597\,\text{m}^2/\text{m}$ of outer surface per tube, the tube length is
$108/(118\times 0.0597) = 15.3\,\text{m}$, which exceeds a standard tube and would
force multiple tube lengths in series or a longer non-standard tube. The velocity
target and the area target pull against each other through the tube count, and the
designer trades them: the high-velocity design is a long, thin, high-film,
high-pressure-drop bundle, the low-velocity design a short, fat, low-film,
low-pressure-drop bundle, both meeting the same area. The case study chose the
short fat bundle and accepted the low film; this example shows the long thin
alternative and why the standard tube length caps how far the velocity target can
be pushed.

\textbf{The arithmetic mean and the log-mean compared.} The derivation showed the
arithmetic mean of the two end differences always exceeds the log-mean and that
sizing on the arithmetic mean undersizes the exchanger. The size of the error is
worth a number. Take three duties with the same hot-end difference $\theta_1 =
40\,\si{\kelvin}$ and a falling cold-end difference. With $\theta_2 = 30\,
\si{\kelvin}$ the arithmetic mean is $35.0$ and the log-mean is $(40 - 30)/\ln(40/
30) = 34.8\,\si{\kelvin}$, a $0.6\,\%$ gap. With $\theta_2 = 10\,\si{\kelvin}$ the
arithmetic mean is $25.0$ and the log-mean is $(40 - 10)/\ln(4) = 21.6\,\si{
\kelvin}$, a $16\,\%$ gap that would undersize the area by the same fraction. With
$\theta_2 = 2\,\si{\kelvin}$ the arithmetic mean is $21.0$ while the log-mean is
$(40 - 2)/\ln(20) = 12.7\,\si{\kelvin}$, a $65\,\%$ over-statement that would size
the exchanger at little more than half the area it needs. The rule the numbers
support is the working one: the arithmetic mean is acceptable only when the two end
differences are within a factor of about two, which is the $\theta_2 = 30$ case;
once an exchanger makes a close approach at one end, the cold-end difference
collapses, the two means diverge, and only the log-mean sizes the area correctly.
The same collapse is what makes close-approach condensers and regenerators
area-hungry, and it is read here directly off the mean.

\textbf{Rating a one-shell-pass unit with the closed-form relation.} The
one-shell-pass effectiveness derived above can be exercised on a concrete rating.
A one-shell-pass, two-tube-pass exchanger has $UA = 24{,}000\,\text{W}/\si{
\kelvin}$ and cools a process stream with $C_{\min} = 12{,}000\,\si{\watt\per
\kelvin}$ against a coolant with $C_{\max} = 20{,}000\,\si{\watt\per\kelvin}$, so
$\text{NTU} = 24{,}000/12{,}000 = 2.0$ and $C_r = 0.6$. The blend rate is $\gamma =
\sqrt{1 + 0.6^2} = \sqrt{1.36} = 1.166$, so $\text{NTU}\,\gamma = 2.33$ and
$e^{-2.33} = 0.0973$. The bracketed group is $(1 + 0.0973)/(1 - 0.0973) = 1.0973/
0.9027 = 1.216$, so
\[
\varepsilon_1 = \frac{2}{(1 + 0.6) + 1.166\times 1.216}
= \frac{2}{1.6 + 1.418} = \frac{2}{3.018} = 0.663.
\]
For comparison the pure counter-flow relation at the same NTU and $C_r$ gives
$\varepsilon = (1 - e^{-2.0\times 0.4})/(1 - 0.6\,e^{-2.0\times 0.4}) = (1 -
0.449)/(1 - 0.270) = 0.551/0.730 = 0.755$, and the parallel-flow relation gives
$(1 - e^{-2.0\times 1.6})/1.6 = (1 - 0.0408)/1.6 = 0.600$. The shell-and-tube unit
sits at $0.663$, between the $0.600$ parallel-flow floor and the $0.755$
counter-flow ceiling, exactly as the arrangement figure showed. The duty is
$\dot{Q} = 0.663\times 12{,}000\times \Delta T_{\max}$, so an engineer who had
carelessly used the counter-flow relation would have over-stated the duty by
$0.755/0.663 = 14\,\%$, the size of the routing error the closed-form relation
exists to prevent.

\textbf{Closing the LMTD and NTU routes on the same shell-and-tube unit.} The
equivalence proof asserted the two methods are one equation; the previous example
gives the numbers to check it on a non-trivial arrangement. The unit above runs at
$\varepsilon_1 = 0.663$, $C_{\min} = 12{,}000$, $\Delta T_{\max} = 100\,\si{
\kelvin}$ say, so $\dot{Q} = 0.663\times 12{,}000\times 100 = 796\,\text{kW}$. The
outlets follow: the process stream cools by $\varepsilon_1\Delta T_{\max} = 66.3\,
\si{\kelvin}$ to $T_{h,o} = T_{h,i} - 66.3$, and the coolant warms by
$\varepsilon_1 C_r\Delta T_{\max} = 39.8\,\si{\kelvin}$. With $T_{h,i} = 130\,
\degC$ and $T_{c,i} = 30\,\degC$ the four temperatures are $T_{h,i} = 130$,
$T_{h,o} = 63.7$, $T_{c,i} = 30$, $T_{c,o} = 69.8\,\degC$. The counter-flow
terminal differences are $\theta_1 = 130 - 69.8 = 60.2$ and $\theta_2 = 63.7 - 30
= 33.7\,\si{\kelvin}$, so $\Delta T_{\text{LMTD,CF}} = (60.2 - 33.7)/\ln(60.2/
33.7) = 26.5/0.580 = 45.7\,\si{\kelvin}$. The LMTD route then needs $UAF = \dot{Q}
/\Delta T_{\text{LMTD,CF}} = 796{,}000/45.7 = 17{,}420\,\text{W}/\si{\kelvin}$, and
since the NTU route built the unit at $UA = 24{,}000$, the implied correction
factor is $F = 17{,}420/24{,}000 = 0.726$. That $F$ falls just below the $0.75$
floor, which is the same warning the effectiveness already gave: at $\text{NTU} =
2.0$ and $C_r = 0.6$ the one-shell-pass unit is being pushed close to what a single
shell can reach, and either a second shell or a relaxed approach is wanted. The two
methods agree on the physics and disagree on nothing; the LMTD route surfaces the
marginal $F$ that the effectiveness route had folded silently into $\varepsilon_1$.

\textbf{A near-balanced regenerator and the cost of the last few percent.} The
pasteurizer case asserted that lifting the regeneration ratio from $0.90$ to
$0.95$ more than doubles the plate count; the balanced-flow relation makes the
sensitivity exact and general. For a balanced ($C_r = 1$) counter-flow section the
effectiveness is $\varepsilon = \text{NTU}/(1 + \text{NTU})$, which inverts to
$\text{NTU} = \varepsilon/(1 - \varepsilon)$. The area is proportional to NTU at
fixed $U$ and $C$, so the area to reach effectiveness $\varepsilon$ scales as
$\varepsilon/(1 - \varepsilon)$, a function that runs away as $\varepsilon \to 1$.
Tabulating: $\varepsilon = 0.80$ needs $\text{NTU} = 4.0$; $\varepsilon = 0.90$
needs $9.0$; $\varepsilon = 0.95$ needs $19.0$; $\varepsilon = 0.98$ needs $49.0$.
The area to go from $0.90$ to $0.95$ is $19/9 = 2.1$ times the area for $0.90$
alone, and the area to reach $0.98$ is more than five times. The marginal area to
buy the next point of effectiveness is $d(\text{NTU})/d\varepsilon = 1/(1 -
\varepsilon)^2$, which at $\varepsilon = 0.95$ is $400$ transfer units per unit
effectiveness against $100$ at $\varepsilon = 0.90$. The balanced regenerator is
the worst case for this runaway because $C_r = 1$ gives the flattest effectiveness
curve, and it is exactly the case dairy and ventilation heat-recovery units live
in, which is why their design settles where the marginal plate stops earning its
recovered heat rather than at the highest reachable effectiveness.

\textbf{Stream splitting to hit two outlet targets at once.} A single coolant
stream sometimes serves two duties that want different approaches, and the
multi-stream balance fixes how to split it. Suppose $C_c = 30{,}000\,\si{\watt\per
\kelvin}$ of cooling water must cool two process streams, the first ($C_1 =
10{,}000\,\si{\watt\per\kelvin}$, in at $120\,\degC$) to be taken to $60\,\degC$
and the second ($C_2 = 8000\,\si{\watt\per\kelvin}$, in at $90\,\degC$) to $50\,
\degC$. The two duties are $\dot{Q}_1 = 10{,}000\times 60 = 600\,\text{kW}$ and
$\dot{Q}_2 = 8000\times 40 = 320\,\text{kW}$. If the water (in at $30\,\degC$) is
split into two parallel branches, each branch's outlet follows from its branch
balance, and the recombined return temperature is the flow-weighted mix. Sending a
fraction $f$ of the water to the first duty, branch one warms by $\dot{Q}_1/(f
C_c)$ and branch two by $\dot{Q}_2/((1 - f)C_c)$. To return both branches at a
common temperature, set the two outlets equal: $30 + 600{,}000/(f\times 30{,}000) =
30 + 320{,}000/((1 - f)\times 30{,}000)$, which gives $600/f = 320/(1 - f)$, so
$600(1 - f) = 320 f$, $600 = 920 f$, and $f = 0.652$. The first duty takes
$65\,\%$ of the water and both branches return at $30 + 600{,}000/(0.652\times
30{,}000) = 30 + 30.7 = 60.7\,\degC$. The split that matches the branch outlets is
the one that avoids mixing streams at different temperatures, which would generate
avoidable entropy at the junction, the network-level form of the second-law
argument the counter-flow-versus-parallel-flow exercise made for a single unit.

\textbf{Zoning a de-superheating condenser with subcooling.} A condenser fed
superheated vapour and asked to deliver subcooled liquid cannot be sized on one
log-mean and one coefficient, because the process side passes through three regimes
with three different overall coefficients, and treating it as a single zone
mis-sizes the area. Take a vapour stream that enters at $130\,\degC$ superheated,
condenses at a saturation temperature of $80\,\degC$, and leaves as liquid
subcooled to $55\,\degC$, against cooling water rising from $30\,\degC$ to
$60\,\degC$. Split the duty into three zones and size each on its own driving
difference. Suppose the desuperheating zone carries $12\,\%$ of the total duty, the
condensing zone $73\,\%$, and the subcooling zone $15\,\%$; for a total duty of
$\dot{Q} = 3.0\,\text{MW}$ these are $0.36$, $2.19$, and $0.45\,\text{MW}$. The
water splits its rise in proportion to the duty carried in each zone, so it warms
by $0.12\times 30 = 3.6\,\si{\kelvin}$ across the subcooling zone at the cold end,
by $0.73\times 30 = 21.9\,\si{\kelvin}$ across the condensing zone, and by
$0.15\times 30 = 4.5\,\si{\kelvin}$ across the desuperheating zone at the hot end,
so the water temperatures at the two zone boundaries are $30 + 3.6 = 33.6\,\degC$
and $33.6 + 21.9 = 55.5\,\degC$. The condensing zone then runs between
$\theta_1 = 80 - 55.5 = 24.5$ and $\theta_2 = 80 - 33.6 = 46.4\,\si{\kelvin}$ for a
log-mean of $(46.4 - 24.5)/\ln(46.4/24.5) = 34.3\,\si{\kelvin}$; at a condensing
coefficient of $U = 2500\,\si{\watt\per\meter\squared\per\kelvin}$ the condensing
area is $2.19\times 10^6/(2500\times 34.3) = 25.5\,\text{m}^2$. The two single-phase
zones run lower coefficients, say $U = 900$ for the vapour desuperheat and $U =
1400$ for the liquid subcool, and their smaller duties across their own log-means
add perhaps $8$ and $6\,\text{m}^2$, so the zoned total lands near
$40\,\text{m}^2$. A single-zone estimate on the overall terminal temperatures and
one blended coefficient would miss this by ten to twenty percent because it smears
the high condensing coefficient across the whole area and the low single-phase
coefficients get no separate accounting. Figure
\ref{fig:vol04ch06:condenser-zones} draws the three-zone profile and the flat
condensing plateau that carries most of the duty. The lesson is the one the LMTD
derivation flagged: the log-mean holds only where $U$ is constant, so a unit whose
$U$ jumps at a phase boundary is sectioned into zones within which the assumption
holds, and the zones are chained.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-condenser-zones.tex}

\textbf{Reading a fouling resistance off a Wilson plot.} An operating exchanger's
overall coefficient can be split into its tube-side film and the rest without
opening the unit, by running it at several tube-side velocities and plotting the
measured resistance against the velocity group the film follows. In turbulent flow
the tube-side film obeys $h_i \propto u^{0.8}$, so the tube-side resistance referred
out is $C_1 u^{-0.8}$ for some constant $C_1$, while the wall, the fouling, and the
shell-side film do not move with tube-side velocity and sum to a constant $C_0$.
The measured overall resistance is therefore $1/U_o = C_0 + C_1 u^{-0.8}$, a
straight line in the variable $x = u^{-0.8}$. Take a clean unit measured at four
velocities, giving the fitted line $1/U_o = 0.55 + 1.30\,x$ in units of $10^{-3}\,
\text{m}^2\text{K}/\text{W}$: the intercept $0.55\times 10^{-3}$ is the sum of the
velocity-independent resistances and the slope $1.30\times 10^{-3}$ is the
tube-side film group. At a design velocity of $u = 1.5\,\text{m/s}$ the group is
$u^{-0.8} = 1.5^{-0.8} = 0.73$, so the clean resistance is $0.55 + 1.30\times 0.73
= 1.50\times 10^{-3}$ and $U_o = 667\,\si{\watt\per\meter\squared\per\kelvin}$.
Repeat the campaign a year later and the fitted line shifts to $1/U_o = 0.95 +
1.30\,x$: the slope is unchanged, so the tube-side film is unchanged, but the
intercept has risen by $0.40\times 10^{-3}\,\text{m}^2\text{K}/\text{W}$, which is
the fouling resistance that has grown, measured directly in the field. Figure
\ref{fig:vol04ch06:wilson-plot} shows the construction and the parallel-line shift.
The Wilson plot is the working tool for separating a film that scales with velocity
from a fouling that does not, and its parallel-line signature is the cleanest field
evidence that a decline is fouling rather than a change in the process film.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-wilson-plot.tex}

\textbf{The critical radius of insulation on a small tube.} Adding insulation to a
pipe does not always cut the heat loss, and the exception is worth a number because
it catches designers who reason that more insulation is always better. Insulation
adds a conduction resistance that grows with the outer radius, but it also grows the
outer surface area, which cuts the convective resistance at the outside, and on a
small enough tube the second effect wins until a critical radius is passed. For a
bare cylinder of outer radius $r$ carrying insulation of conductivity $k$ to an
outer radius $r_o$, with an outside film coefficient $h$, the two outer resistances
per unit length are the insulation conduction $\ln(r_o/r)/(2\pi k)$ and the surface
convection $1/(2\pi r_o h)$. Their sum has a minimum where $d/dr_o$ vanishes, at the
\engterm{critical radius} $r_{\text{crit}} = k/h$. Take a small tube carrying an
electrical cable insulated with a polymer of $k = 0.15\,\si{\watt\per\meter\per
\kelvin}$ in still air with $h = 8\,\si{\watt\per\meter\squared\per\kelvin}$: the
critical radius is $r_{\text{crit}} = 0.15/8 = 0.0188\,\text{m}$, about
$19\,\si{\milli\meter}$. A cable of outer radius $5\,\si{\milli\meter}$ insulated
up to less than $19\,\si{\milli\meter}$ therefore sheds \emph{more} heat than the
bare cable, which is the desirable behaviour for a current-carrying wire that must
stay cool, and the perverse behaviour for a hot pipe that is meant to be insulated.
For the district-heating main of the estimation exercise the numbers run the other
way: with $k = 0.025$ and $h$ of order $10$, the critical radius is
$2.5\,\si{\milli\meter}$, far inside the $100\,\si{\milli\meter}$ pipe radius, so
every millimetre of insulation cuts the loss and the intuition holds. The critical
radius is a small-tube effect that vanishes as soon as the bare radius exceeds
$k/h$, which is why it matters for wires and capillaries and never for process
piping.

\textbf{A two-fluid balance that pins the third temperature.} An economiser or a
recuperator often has three of its four terminal temperatures fixed by the process
and the fourth left to the balance, and getting the balance right is what closes the
duty. Suppose a gas turbine recuperator preheats $6\,\text{kg/s}$ of compressor
discharge air ($c_p = 1.05\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$) using
$6.2\,\text{kg/s}$ of turbine exhaust ($c_p = 1.10\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$). The exhaust enters the recuperator at $480\,\degC$, the air enters
at $210\,\degC$, and the recuperator is rated at an air-side effectiveness of
$\varepsilon = 0.75$. The capacity rates are $C_{\text{air}} = 6\times 1050 =
6300\,\si{\watt\per\kelvin}$ and $C_{\text{exh}} = 6.2\times 1100 = 6820\,\si{\watt
\per\kelvin}$, so $C_{\min} = 6300$ (the air) and $C_r = 6300/6820 = 0.92$, a
near-balanced unit. The maximum possible duty is $C_{\min}(T_{\text{exh,i}} -
T_{\text{air,i}}) = 6300\times (480 - 210) = 1.70\,\text{MW}$, so the actual duty is
$\dot{Q} = 0.75\times 1.70 = 1.28\,\text{MW}$. The air, being $C_{\min}$, warms by
the full effectiveness fraction: $\Delta T_{\text{air}} = \varepsilon(T_{\text{exh,i}}
- T_{\text{air,i}}) = 0.75\times 270 = 202\,\si{\kelvin}$, leaving the air at
$210 + 202 = 412\,\degC$ to enter the combustor. The exhaust, being $C_{\max}$,
cools by less: $\Delta T_{\text{exh}} = \dot{Q}/C_{\text{exh}} = 1.28\times 10^6/
6820 = 188\,\si{\kelvin}$, leaving the exhaust at $480 - 188 = 292\,\degC$ to the
stack. The check that the balance closes is that the two duties agree,
$C_{\text{air}}\times 202 = 1.27\,\text{MW}$ against $C_{\text{exh}}\times 188 =
1.28\,\text{MW}$, within rounding. The recovered $1.28\,\text{MW}$ is heat the
combustor no longer burns fuel to supply, and because the unit is near-balanced its
effectiveness sits on the flat part of the curve where the next point of recovery
is expensive in area, the same regenerator economics the pasteurizer showed at
$C_r = 1$.

\textbf{Sizing a plate exchanger from the channel count and the target film.} The
plate unit is sized the other way from the shell unit: the channel count is chosen
to hold a target channel velocity that keeps the film high, and the plate count then
follows from the transfer units the duty needs. Take a district-heating substation
exchanger passing $4\,\text{kg/s}$ of network water ($\rho = 980\,\text{kg/m}^3$,
$\mu = 4.5\times 10^{-4}\,\text{Pa}\cdot\text{s}$) on each side, with a channel gap
$b = 2.5\,\si{\milli\meter}$ and a plate width $w = 0.4\,\text{m}$. A target channel
velocity of $0.4\,\text{m/s}$ needs a flow area per side of $\dot{m}/(\rho u) =
4/(980\times 0.4) = 0.0102\,\text{m}^2$, and each channel offers $b\,w = 0.0025
\times 0.4 = 0.001\,\text{m}^2$, so the channels per side are $0.0102/0.001 = 10.2$,
round to $10$. That fixes the channel Reynolds number at $Re = \rho u d_h/\mu$ with
$d_h = 2b = 5\,\si{\milli\meter}$: $Re = 980\times 0.4\times 0.005/(4.5\times
10^{-4}) = 4360$, turbulent-enough for the corrugation to hold a high film. A plate
Nusselt correlation of the form $Nu = 0.3\,Re^{0.65}Pr^{0.4}$ with $Pr \approx 3$
gives $Nu = 0.3\times 4360^{0.65}\times 3^{0.4} = 0.3\times 262\times 1.55 = 122$,
so the water-side film is $h = Nu\,k/d_h = 122\times 0.63/0.005 = 15{,}400\,\si{\watt
\per\meter\squared\per\kelvin}$, and with the same film on both sides and a thin
steel plate the overall coefficient is near $6500\,\si{\watt\per\meter\squared\per
\kelvin}$. If the duty needs $\text{NTU} = 3$ on a balanced section with $C =
\dot{m}c_p = 4\times 4180 = 16{,}720\,\si{\watt\per\kelvin}$, the area is
$\text{NTU}\,C/U = 3\times 16{,}720/6500 = 7.7\,\text{m}^2$, which at
$0.16\,\text{m}^2$ of transfer area per thermal plate is about $48$ thermal plates.
The channel count set the velocity and the film; the plate count then met the
transfer units, and the two are checked against each other rather than either being
chosen alone.

\textbf{The pinch inside a single exchanger and the minimum approach.} A close
approach is what makes an exchanger area-hungry, and the sensitivity is worth a
number because it governs where a design settles. Take a fixed duty of $\dot{Q} =
500\,\text{kW}$ cooling a stream against a coolant, with the hot-end difference held
at $\theta_1 = 40\,\si{\kelvin}$ and the cold-end difference $\theta_2$ chosen by
the approach the designer accepts, at $U = 1000\,\si{\watt\per\meter\squared\per
\kelvin}$. At $\theta_2 = 20\,\si{\kelvin}$ the log-mean is $(40 - 20)/\ln(2) =
28.9\,\si{\kelvin}$ and the area is $500{,}000/(1000\times 28.9) = 17.3\,\text{m}^2$.
Tighten the approach to $\theta_2 = 10\,\si{\kelvin}$ and the log-mean falls to
$(40 - 10)/\ln(4) = 21.6\,\si{\kelvin}$, so the area rises to $23.1\,\text{m}^2$.
Tighten again to $\theta_2 = 5\,\si{\kelvin}$ and the log-mean is $(40 - 5)/\ln(8) =
16.8\,\si{\kelvin}$, area $29.7\,\text{m}^2$; at $\theta_2 = 2\,\si{\kelvin}$ the
log-mean is $(40 - 2)/\ln(20) = 12.7\,\si{\kelvin}$, area $39.4\,\text{m}^2$. Halving
the approach from $20$ to $10\,\si{\kelvin}$ cost a third more area; halving it again
to $5$ cost another third; the marginal area to buy the next kelvin of approach
climbs without bound as $\theta_2 \to 0$. This single-exchanger sensitivity is the
same trade the pinch construction reads across a whole network: the minimum approach
$\Delta T_{\min}$ chosen in the composite-curve construction sets the recovered heat
against the area, and it settles where the marginal area cost meets the marginal
utility saving, exactly as the cold-end approach settles the size of one unit.

\textbf{Sizing a filmwise condenser and reading the film penalty.} A condenser's
coefficient is capped by the liquid film that drains down its tubes, and the
Nusselt filmwise theory gives the number. Take saturated steam at $T_{\text{sat}}
= 60\,\degC$ condensing on the outside of a horizontal tube of outer diameter
$d_o = 25\,\si{\milli\meter}$ held at a wall temperature $T_w = 50\,\degC$, so the
film drives across $\Delta T = 10\,\si{\kelvin}$. The Nusselt result for a single
horizontal tube is
\[
h = 0.729\left[\frac{\rho_\ell(\rho_\ell - \rho_v) g\, h_{fg}\, k_\ell^3}
{\mu_\ell\,\Delta T\, d_o}\right]^{1/4},
\]
and with condensate properties near $55\,\degC$ ($\rho_\ell = 986\,\text{kg/m}^3$,
$k_\ell = 0.65\,\si{\watt\per\meter\per\kelvin}$, $\mu_\ell = 5.0\times 10^{-4}\,
\text{Pa}\cdot\text{s}$, $h_{fg} = 2.36\times 10^6\,\text{J/kg}$, $\rho_v$
negligible) the bracket is
\[
\frac{986\times 986\times 9.81\times 2.36\times 10^6\times 0.65^3}
{5.0\times 10^{-4}\times 10\times 0.025} = \frac{2.49\times 10^{13}}
{1.25\times 10^{-4}} = 1.99\times 10^{17},
\]
whose fourth root is $2.11\times 10^4$, so $h = 0.729\times 2.11\times 10^4 =
1.54\times 10^4\,\si{\watt\per\meter\squared\per\kelvin}$. A filmwise coefficient
of about $15{,}000$ is large against a liquid single-phase film, which is why the
condensing side rarely governs, but it is far below the dropwise mode, where a
non-wetting surface would raise the coefficient several-fold by keeping stretches
of bare metal exposed. Figure \ref{fig:vol04ch06:condensation-modes} contrasts the
two. The filmwise value is the one to design on, because the surface treatments
that sustain dropwise condensation wear off in service and the condenser settles
into the filmwise mode the Nusselt theory predicts. Stacking the tubes in a
vertical column lowers the coefficient further, because condensate from the upper
tubes floods the lower ones, thickening their films, an effect the design accounts
for with a row-count correction below unity.

\textbf{The pumping fraction that turns a good exchanger uneconomic.} An exchanger
earns its keep only if the pumping power it costs is small against the heat it
moves, and the ratio is worth a number because it fixes the velocity ceiling. Take
a liquid stream of density $\rho = 900\,\text{kg/m}^3$ and specific heat
$c_p = 2200\,\text{J}/(\text{kg}\cdot\si{\kelvin})$ cooled through $\Delta T =
40\,\si{\kelvin}$, so each kilogram carries $\dot{q}/\dot{m} = c_p\Delta T =
2200\times 40 = 88{,}000\,\text{J/kg}$ of duty. Push it down a tube bundle at
$u = 1.5\,\text{m/s}$ against a total pressure drop of $\Delta p = 40\,\text{kPa}$;
the pumping work per kilogram is $\Delta p/(\rho\,\eta_p)$ at a pump efficiency
$\eta_p = 0.65$, which is $40{,}000/(900\times 0.65) = 68\,\text{J/kg}$. The
pumping fraction is $68/88{,}000 = 7.7\times 10^{-4}$, under a tenth of a percent,
comfortably inside the customary band. Now double the velocity to $3\,\text{m/s}$:
the pressure drop rises as $u^2$ to $160\,\text{kPa}$, the pumping work per
kilogram to $273\,\text{J/kg}$, and the fraction to $3.1\times 10^{-3}$, still
acceptable but four times worse for a film gain of only $2^{0.8} = 1.7$ times.
Figure \ref{fig:vol04ch06:pumping-fraction} plots the fraction against velocity
and marks the band. The lesson the fraction makes plain is that the pumping cost of
a heat exchanger is a dimensionless efficiency, the ratio of the work spent moving
the fluid to the heat the fluid carries, and it climbs as about $u^{2.2}$, so the
velocity ceiling is reached when moving the fluid costs a noticeable fraction of
the heat delivered, long before the pump itself runs out of head.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-pumping-fraction.tex}

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-condensation-modes.tex}

\textbf{The parallel-flow ceiling and the duty a routing forfeits.} The
effectiveness ceiling of parallel-flow was derived above as $1/(1 + C_r)$; a
concrete pair of duties shows what that ceiling costs. Take two exchangers at
$C_r = 0.5$, one counter-flow and one parallel-flow, both fed a maximum driving
difference of $\Delta T_{\max} = 80\,\si{\kelvin}$ with $C_{\min} = 5000\,\si{\watt
\per\kelvin}$, and ask what each delivers at a large $\text{NTU} = 4$. The
counter-flow relation gives $\varepsilon = (1 - e^{-4\times 0.5})/(1 - 0.5\,e^{-4
\times 0.5}) = (1 - 0.135)/(1 - 0.0677) = 0.865/0.932 = 0.928$, a duty of
$0.928\times 5000\times 80 = 371\,\text{kW}$. The parallel-flow relation gives
$\varepsilon = (1 - e^{-4\times 1.5})/1.5 = (1 - 0.00248)/1.5 = 0.665$, a duty of
$0.665\times 5000\times 80 = 266\,\text{kW}$. The parallel unit delivers
$105\,\text{kW}$ less on the same area, and buying more area cannot close the gap:
the parallel ceiling at $C_r = 0.5$ is $1/(1 + 0.5) = 0.667$, which the unit has
already all but reached at $\text{NTU} = 4$, so doubling the area would lift it by
under a fifth of a percent. Figure \ref{fig:vol04ch06:effectiveness-ceiling} draws
the two ceilings against $C_r$, and the shaded gap between them is exactly this
forfeited duty read across all capacity-rate ratios. The lesson is that the routing
choice is not a small correction but a hard ceiling: a parallel-flow unit at finite
$C_r$ leaves duty on the table that no amount of metal recovers, which is why
counter-flow is the default and parallel-flow is chosen only when a wall-temperature
or thermal-stress constraint forces the co-current arrangement.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-effectiveness-ceiling.tex}

\textbf{A run-around coil loop between two distant streams.} When a hot stream and
a cold stream are too far apart to bring into one exchanger, a pumped liquid loop
carries heat between two coils, one in each duct, and the loop's effectiveness is
the product of two exchanger effectivenesses in series through the circulating
liquid. Take a hot exhaust duct and a cold intake duct, each handling $4\,\text{kg/s}$
of air ($C_{\text{air}} = 4\times 1005 = 4020\,\si{\watt\per\kelvin}$), connected by
a water loop circulating $\dot{m}_w = 2\,\text{kg/s}$ ($C_w = 2\times 4180 =
8360\,\si{\watt\per\kelvin}$). Each coil is a cross-flow air-to-water unit with
$UA = 6000\,\text{W}/\si{\kelvin}$, so each has $\text{NTU} = UA/C_{\min} =
6000/4020 = 1.49$ on the air side with $C_r = 4020/8360 = 0.48$, giving a per-coil
effectiveness near $\varepsilon_{\text{coil}} = 0.68$ from the cross-flow
correlation. The loop effectiveness is bounded by the coils in series and by the
water flow that shuttles the heat; a standard result for a balanced run-around loop
is $1/\varepsilon_{\text{loop}} = 1/\varepsilon_1 + 1/\varepsilon_2 - 1$, which for
two equal coils gives $\varepsilon_{\text{loop}} = \varepsilon_{\text{coil}}/
(2 - \varepsilon_{\text{coil}}) = 0.68/(2 - 0.68) = 0.52$. The loop recovers a
lower effectiveness than either coil alone because the heat crosses two
temperature approaches instead of one, and the recovered duty is $\dot{Q} =
\varepsilon_{\text{loop}} C_{\text{air}}(T_{h,i} - T_{c,i})$. The run-around loop
is the answer when carry-over rules out the rotary wheel and distance rules out a
direct exchanger, and its price is the doubled approach that halves the
effectiveness the wheel would have reached.

\textbf{Trimming a design with a bypass to hit a variable target.} A process that
needs a controlled outlet over a range of loads often oversizes the exchanger and
bypasses part of the flow to trim the outlet, and the bypass fraction follows from
the mixing balance run backward. Suppose an exchanger cools a process stream from
$120\,\degC$ and, at full flow through the bundle, would deliver $55\,\degC$, but
the downstream unit wants a controllable $70\,\degC$. Bypassing a fraction $f$ of
the hot feed around the bundle and remixing gives $T_{\text{mix}} = f\times 120 +
(1 - f)\times 55$, set equal to the target: $70 = 120 f + 55(1 - f) = 55 + 65 f$,
so $f = 15/65 = 0.23$. Routing $23\,\%$ of the stream around the bundle lifts the
mixed outlet from $55$ to $70\,\degC$, and closing the bypass valve toward zero
drops the outlet back toward $55\,\degC$ as the load rises, which is the control
handle the oversized exchanger was built to provide. The same arithmetic that
diagnosed an unwanted bypass earlier is here run as a design tool: a deliberate,
measured bypass is a control element, while an unmeasured one is a fault, and the
difference is whether the mixing balance is being used on purpose or discovered
after the outlet drifts.

\textbf{A temperature cross that no single shell can reach.} A duty asks the cold
outlet to rise above the hot outlet, a temperature cross, and reading when a cross
appears is what tells a designer to add a shell before wasting effort on a single
one. Take a duty cooling a hot stream from $T_{h,i} = 150\,\degC$ to
$T_{h,o} = 60\,\degC$ while heating a cold stream from $T_{c,i} = 40\,\degC$ to
$T_{c,o} = 90\,\degC$. The cold outlet at $90\,\degC$ exceeds the hot outlet at
$60\,\degC$ by $30\,\si{\kelvin}$, so the streams cross, which a pure counter-flow
unit handles but a one-shell-pass, two-tube-pass unit cannot, because half its tube
passes run co-current and cannot push the cold stream past the hot outlet. The
correction-factor parameters are $P = (90 - 40)/(150 - 40) = 50/110 = 0.45$ and
$R = (150 - 60)/(90 - 40) = 90/50 = 1.8$, and reading the one-shell-pass chart at
$P = 0.45$, $R = 1.8$ lands on the steep collapse where $F$ falls below the
$0.75$ floor, the chart's way of announcing the cross. The repair is a
two-shell-pass unit, which restores enough counter-flow character to lift $F$ back
above the floor: splitting the duty across two shells in series gives each shell a
smaller $P$ that sits on the flat part of its curve. The lesson is to compute $P$
and $R$ and check for a cross before sizing: a cross with a single shell is not a
small $F$ penalty but an impossibility, and the chart's collapse is the signal to
change the configuration rather than enlarge it.

\textbf{The wall temperature and the dew-point margin on a cold surface.} An
economiser or an air heater cooling a flue gas must keep its cold-end metal above
the acid dew point, and the wall temperature follows from the series network read
at the cold end. The metal sits between the two films, so its temperature is the
gas temperature pulled toward the coolant temperature by the ratio of the two
film resistances: $T_w = T_{\text{gas}} - (T_{\text{gas}} - T_{\text{cool}})
\times R_{\text{gas}}/(R_{\text{gas}} + R_{\text{cool}})$, where $R = 1/h$ on each
side. Take a cold-end gas at $T_{\text{gas}} = 130\,\degC$ against feedwater at
$T_{\text{cool}} = 105\,\degC$, with a gas-side film $h_g = 60\,\si{\watt\per\meter
\squared\per\kelvin}$ and a water-side film $h_w = 3000\,\si{\watt\per\meter
\squared\per\kelvin}$. The resistances are $R_g = 1/60 = 0.0167$ and
$R_w = 1/3000 = 0.00033\,\text{m}^2\cdot\si{\kelvin}/\text{W}$, so the wall sits at
$T_w = 130 - (130 - 105)\times 0.0167/(0.0167 + 0.00033) = 130 - 25\times 0.980 =
130 - 24.5 = 105.5\,\degC$. The metal runs within a kelvin of the water, not of the
gas, because the water film is fifty times tighter, so the cold-end metal is at
$105.5\,\degC$, and if the acid dew point for this sulphur-bearing fuel is
$120\,\degC$ the metal is below it and will corrode. The fix reads straight off the
balance: raise the feedwater inlet temperature, which lifts the metal one-for-one,
or accept a corrosion-resistant cold-end material. The general lesson is that a
wall sits at the temperature of whichever side has the tighter film, which is why a
cold surface against a strong liquid film runs cold and dew-point corrosion is a
cold-end problem set by the coolant, not the gas.

\textbf{A transient warm-up and the exchanger's thermal time constant.} A heat
exchanger does not reach its steady outlet the instant flow starts; its metal must
first warm to its running temperature, and the lag follows a first-order time
constant worth estimating for a control loop. Model the exchanger metal as a lump
of mass $m_{\text{metal}} = 400\,\text{kg}$ and specific heat $c_{\text{metal}} =
480\,\text{J}/(\text{kg}\cdot\si{\kelvin})$ (carbon steel), heated by a process
stream through an effective conductance $UA = 20{,}000\,\text{W}/\si{\kelvin}$. The
thermal time constant is $\tau = m_{\text{metal}} c_{\text{metal}}/(UA) = 400\times
480/20{,}000 = 9.6\,\text{s}$, so the metal, and with it the outlet, reaches
$63\,\%$ of its final temperature step in about ten seconds and settles within a
few percent after three time constants, roughly half a minute. A control loop
tuned faster than this lag will hunt, because the exchanger cannot follow a
setpoint change faster than its metal can warm, and a loop tuned much slower wastes
the exchanger's responsiveness. The estimate scales with the metal mass, so a
massive shell unit lags minutes while a thin-plate exchanger lags seconds, which is
one reason plate units are favoured where fast thermal response matters, as in the
pasteurizer whose hold temperature must not droop when the feed rate steps.

\textbf{Choosing between two candidate exchangers on annualised cost.} A design
decision between two units that both meet the duty is settled by annualised cost,
the capital amortised plus the running pumping cost, and the arithmetic is worth
one pass. Exchanger A is a compact high-velocity unit costing \$40{,}000 with a
pumping power of $6\,\text{kW}$; exchanger B is a larger low-velocity unit costing
\$60{,}000 with a pumping power of $1.5\,\text{kW}$. Amortise the capital over ten
years at a simple annual charge of a tenth of the capital, so A charges
\$4{,}000/yr and B charges \$6{,}000/yr in capital. The pumping runs $8000\,\text{h/yr}$
at an electricity price of \$0.10/kWh, so A costs $6\times 8000\times 0.10 =
\$4{,}800$/yr and B costs $1.5\times 8000\times 0.10 = \$1{,}200$/yr in power. The
annualised totals are A at $4{,}000 + 4{,}800 = \$8{,}800$/yr and B at
$6{,}000 + 1{,}200 = \$7{,}200$/yr, so the larger, more expensive exchanger B is the
cheaper choice over its life because its pumping saving outruns its capital
premium. The result flips if electricity is cheap or the plant runs few hours: at
\$0.03/kWh the power costs fall to \$1{,}440 and \$360, the totals become A at
\$5{,}440 and B at \$6{,}360, and the compact unit wins. The lesson is that the
capital-versus-pumping trade is decided by the operating hours and the energy
price, not by the purchase cost alone, which is why the same duty is met by a
compact unit in a lightly loaded plant and a generous unit in one that runs flat
out.

\textbf{Reading an exchanger's duty from a plant heat balance.} An operating
exchanger's actual duty is often not measured directly but inferred from the
temperatures and flows a plant already logs, and closing the balance both ways is
the check that the readings are trustworthy. Suppose a logged exchanger shows a hot
stream of $\dot{m}_h = 7\,\text{kg/s}$ ($c_{p,h} = 2.3\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$) entering at $110\,\degC$ and leaving at $65\,\degC$, and a cold
stream of $\dot{m}_c = 9\,\text{kg/s}$ ($c_{p,c} = 4.18\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$) entering at $30\,\degC$. The hot-side duty is $\dot{Q}_h =
7\times 2300\times (110 - 65) = 725\,\text{kW}$, which predicts a cold outlet of
$30 + 725{,}000/(9\times 4180) = 30 + 19.3 = 49.3\,\degC$. If the logged cold
outlet reads $49\,\degC$ the balance closes to within the instrument error and the
duty is trustworthy; if it reads $44\,\degC$ the cold-side duty is only
$9\times 4180\times 14 = 527\,\text{kW}$, a $27\,\%$ shortfall against the hot side
that means a heat loss, a bypass, or a mis-calibrated sensor, not a real duty. The
discipline is to compute the duty from both sides and require them to agree: a
closed balance validates the readings and the inferred coefficient $U = \dot{Q}/
(A\,F\,\Delta T_{\text{LMTD}})$, while an open balance flags a measurement or a
process fault before the number is trusted. This is the field counterpart of the
LMTD-versus-NTU closure the methods section ran on paper.

\textbf{The vibration margin of a bundle that passed its thermal check.} A bundle can
meet its duty and its pressure drop and still shake itself apart, so the crossflow
velocity is checked against the shedding resonance before the design is signed. Take a
shell-side crossflow velocity of $u = 1.5\,\text{m/s}$ past tubes of outer diameter
$d_o = 19\,\si{\milli\meter}$ with a Strouhal number $\mathrm{St} = 0.2$. The
vortex-shedding frequency is $f_s = \mathrm{St}\,u/d_o = 0.2\times 1.5/0.019 =
15.8\,\text{Hz}$. The tube's lowest natural frequency for a span between baffles of
$\ell = 0.6\,\text{m}$ follows the beam relation $f_n = (C/2\pi)\sqrt{EI/(m\ell^4)}$;
for a simply-supported span the constant is $C = \pi^2$, and with a steel tube of
$E = 200\,\text{GPa}$, second moment $I = 5.3\times 10^{-9}\,\text{m}^4$, and mass per
length including contained and displaced water $m = 1.1\,\text{kg/m}$, the group is
$EI/(m\ell^4) = 200\times 10^9\times 5.3\times 10^{-9}/(1.1\times 0.6^4) = 1060/0.143
= 7430$, whose square root is $86.2$, so $f_n = (\pi^2/2\pi)\times 86.2 =
(\pi/2)\times 86.2 = 135\,\text{Hz}$. The shedding frequency at $15.8\,\text{Hz}$ is
below a tenth of the $135\,\text{Hz}$ natural frequency, a wide margin, so this
short-span bundle is safe against shedding resonance. The trap is the baffle spacing:
if the span were opened to $\ell = 1.5\,\text{m}$ to cut the shell-side pressure drop,
$f_n$ would fall as $1/\ell^2$ to $135\times(0.6/1.5)^2 = 21.6\,\text{Hz}$, close
enough to the $15.8\,\text{Hz}$ shedding that a modest velocity rise would cross it.
Figure \ref{fig:vol04ch06:vibration-margin} plots the shedding line against velocity
with the natural-frequency band and the fluid-elastic instability ceiling marked. The
lesson the numbers carry is that the baffle spacing is a coupled lever, setting the
span that fixes $f_n$ and the crossflow velocity that fixes $f_s$ at once, so a
pressure-drop fix that opens the baffles can walk a safe bundle into a resonance.

\textbf{Selecting a seawater tube material on life-cycle cost, not purchase price.} A
seawater-cooled condenser must choose its tube metal against pitting and erosion, and
the choice is made on the cost over the plant life, not the cost at purchase. Compare
a copper-nickel (90/10) tube bundle costing \$120{,}000 with an expected tube life of
$12\,\text{years}$ in the site's seawater against a titanium bundle costing
\$260{,}000 with a life beyond the $30\,\text{year}$ plant horizon. Amortise each over
its life with a simple straight-line charge: the copper-nickel bundle charges
$120{,}000/12 = \$10{,}000$/yr and must be replaced twice more over the horizon, while
the titanium bundle charges $260{,}000/30 = \$8{,}700$/yr and never returns to the
shop. The replacement of the copper-nickel bundle also costs two unplanned or planned
outages, each of which idles a power block worth far more per day than the tube metal,
so the lost-production charge dwarfs the tube price: at a conservative
\$50{,}000 per replacement outage spread over the twelve-year life, the copper-nickel
bundle carries a further $\$50{,}000/12 = \$4{,}200$/yr, lifting its annualised total
to $\$14{,}200$/yr against titanium's $\$8{,}700$/yr. Titanium wins decisively once
the outage cost is counted, which is the general finding for severe seawater service
and the reason condensers on coastal plants are titanium despite a purchase price more
than double the copper alloy. The selection reads the resistance stack of the earlier
network the same way, spending capital where the failure mode governs the life, here
pitting and erosion on the seawater side rather than heat-transfer coefficient.

\textbf{A cross-flow unit with one stream mixed.} The cross-flow correlation used so
far assumes both streams unmixed, each holding its transverse temperature profile; an
air cooler whose tube-side fluid is mixed across the pass (a single header) obeys a
different relation, and using the wrong one mis-rates the unit. For cross-flow with
$C_{\min}$ unmixed and $C_{\max}$ mixed the effectiveness is $\varepsilon = 1 -
\exp\{-[1 - e^{-C_r\,\text{NTU}}]/C_r\}$, while for $C_{\min}$ mixed and $C_{\max}$
unmixed it is $\varepsilon = (1/C_r)[1 - \exp\{-C_r(1 - e^{-\text{NTU}})\}]$. Take
$\text{NTU} = 1.5$ and $C_r = 0.5$ and evaluate all three cross-flow forms. Both
unmixed gives $\varepsilon = 1 - \exp\{[\exp(-0.5\times 1.5^{0.78}) - 1]\times
1.5^{0.22}/0.5\}$; with $1.5^{0.78} = 1.36$ and $1.5^{0.22} = 1.093$ the bracket is
$[\exp(-0.68) - 1]\times 1.093/0.5 = (0.507 - 1)\times 2.186 = -1.078$, so
$\varepsilon = 1 - e^{-1.078} = 0.660$. The $C_{\min}$-unmixed, $C_{\max}$-mixed form
gives $\varepsilon = 1 - \exp\{-[1 - e^{-0.5\times 1.5}]/0.5\} = 1 - \exp\{-[1 -
0.472]/0.5\} = 1 - e^{-1.056} = 0.652$. The $C_{\min}$-mixed form gives
$\varepsilon = 2[1 - \exp\{-0.5(1 - e^{-1.5})\}] = 2[1 - \exp\{-0.5\times 0.777\}] =
2[1 - e^{-0.388}] = 2\times 0.322 = 0.643$. The three land within two points of one
another at this NTU, which is the general result that the mixing detail matters little
at modest NTU and $C_r$ but grows at high NTU where the mixed stream's loss of its
profile costs real effectiveness. The working rule is to use the both-unmixed
correlation for finned cross-flow, where neither stream mixes transversely, and to
reach for the mixed forms only when a header genuinely blends one stream, and never to
substitute the counter-flow relation, which here would return $\varepsilon = 0.686$,
overstating every one of the three.

\textbf{The startup transient of a shell unit against a plate unit.} Two exchangers
of equal duty can differ tenfold in how fast they reach it, and the difference is the
metal mass each must warm. Take the shell unit of the transient worked example
($m_{\text{metal}} = 400\,\text{kg}$, $c = 480\,\text{J}/(\text{kg}\cdot\si{\kelvin})$,
$UA = 20{,}000\,\text{W}/\si{\kelvin}$), whose time constant is $\tau =
mc/(UA) = 400\times 480/20{,}000 = 9.6\,\text{s}$. A plate unit of the same duty
holds far less metal, say $m_{\text{metal}} = 40\,\text{kg}$ of thin stainless plate
at $c = 500\,\text{J}/(\text{kg}\cdot\si{\kelvin})$ with a higher $UA = 35{,}000\,
\text{W}/\si{\kelvin}$ from its better coefficient, giving $\tau = 40\times 500/
35{,}000 = 0.57\,\text{s}$. The shell unit reaches $63\%$ of its outlet step in
$9.6\,\text{s}$ and settles after about $29\,\text{s}$; the plate unit reaches the
same fraction in half a second and settles in under two. Figure
\ref{fig:vol04ch06:transient-startup} draws the common first-order approach both
follow, scaled to each unit's own $\tau$. The consequence for control is direct: a
pasteurizer that must hold $72\,\degC$ within a fraction of a kelvin as the feed rate
steps needs the plate unit's fast response, because the shell unit's half-minute lag
would let the hold temperature droop below the legal minimum on every load change
before the metal caught up. The metal mass that buffers a shell unit against upsets is
the same mass that makes it slow, which is the trade the transient time constant
prices.

\textbf{A recovered-heat pinch violation costed as fuel.} An engineer proposes to
raise an economiser's recovery by transferring heat across the process pinch, and the
cost of the violation is worth a number because it looks free on the flowsheet and is
not. Recall the pinch rule that heat carried across the pinch demands an equal amount
of both hot and cold utility to make up for it. Suppose a proposed match carries
$\dot{Q}_{\text{cross}} = 200\,\text{kW}$ across the pinch to close a temperature gap
the composite curves forbid. The rule says this adds $200\,\text{kW}$ of hot utility
above the pinch and $200\,\text{kW}$ of cold utility below it, so the plant burns an
extra $200\,\text{kW}$ of fuel-fired heat and rejects an extra $200\,\text{kW}$ to
cooling for as long as the match runs. Over an $8000\,\text{hour}$ year the wasted hot
utility is $200\times 8000 = 1.6\times 10^6\,\text{kWh} = 1.6\,\text{GWh}$ of heat; at
a boiler fuel cost of \$25/MWh of delivered heat this is $1600\times 25 =
\$40{,}000$/yr thrown away, plus the cooling-utility cost below the pinch. The match
that looked like a saving on the flowsheet is a standing fuel penalty, which is the
quantitative form of the pinch rule against cross-pinch transfer: the composite
curves fix the minimum utilities, and any exchanger placed across the pinch spends
both utilities to move heat the process did not need moved there. The lesson is to
target the utilities with the composite construction before placing matches, so that a
proposed match is checked against the pinch rather than judged by the local
temperature gap it happens to close.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-vibration-margin.tex}

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-transient-startup.tex}

\textbf{Adding a tube pass to lift a marginal correction factor.} When a
one-shell-pass unit lands below the $F = 0.75$ floor, the standard repair is a second
shell in series, and the effectiveness composition makes the lift concrete. Take a
duty that a single one-shell-pass unit rates at $\varepsilon_1 = 0.60$ against a
capacity ratio $C_r = 0.7$, a point the earlier composition formula and the arrangement
figure place just under the counter-flow curve. Two such shells in series, each at
$\varepsilon_1 = 0.60$, compose through the ratio $\Lambda = (1 - \varepsilon_1 C_r)/
(1 - \varepsilon_1) = (1 - 0.60\times 0.7)/(1 - 0.60) = 0.58/0.40 = 1.45$. The
two-shell effectiveness is
\[
\varepsilon = \frac{\Lambda^2 - 1}{\Lambda^2 - C_r}
= \frac{1.45^2 - 1}{1.45^2 - 0.7}
= \frac{2.103 - 1}{2.103 - 0.7} = \frac{1.103}{1.403} = 0.786,
\]
so two shells lift the effectiveness from $0.60$ to $0.786$, and because the counter-flow
ceiling at this NTU and $C_r$ sits near $0.82$, the two-shell arrangement recovers most
of the routing penalty a single shell forfeited. The correction factor rises the same
way: each shell now sees a smaller temperature effectiveness $P$ and sits on the flat
part of its $F$ curve rather than the steep collapse, so both shells run above the
$0.75$ floor where the single shell fell below it. The engineer buys the lift with a
second shell, its nozzles, and its pressure drop, and weighs that against the
alternative of a larger single shell that still cannot cross the single-pass ceiling;
past a duty the single shell cannot reach, the second shell is not an optimisation but
a necessity, which is the reading the marginal $F$ gives.

\textbf{Reading the two goodness factors to rank two candidate surfaces.} The compact
mastery box named the area and volume goodness factors; a two-surface comparison shows
how they decide. Surface A is a plain plate-fin with a Colburn factor $j_A = 0.008$ and
friction factor $f_A = 0.032$ at the design Reynolds number; surface B is an
offset-strip fin with $j_B = 0.011$ and $f_B = 0.055$. The area goodness factor is the
ratio $j/f$: surface A gives $j_A/f_A = 0.008/0.032 = 0.25$ and surface B gives
$0.011/0.055 = 0.20$, so A moves more heat per unit of pumping and wins on the pumping
axis. The compactness pulls the other way: surface B's higher $j$ packs more heat
transfer into a given frontal area, so at equal pumping B needs less volume. The choice
follows the binding constraint the same way the class choice did: a stationary plant
with cheap floor space and dear electricity picks A and accepts the larger core, while
an aircraft or a cryogenic cold box that pays dearly for volume picks B and accepts the
worse $j/f$. The two factors let the ranking be read off published $j$ and $f$ curves
before any geometry is committed, which is how compact-surface selection is actually
done, and the fact that A wins one axis while B wins the other is the general case: a
surface that dominates on both is rare, so the design states which axis binds and reads
the matching factor.

\begin{mastery}
The first-order metal warm-up of the transient examples is a single-lump idealisation,
and the real dynamic response of an exchanger has a second time scale the lump misses,
the transport lag of the fluid travelling the length of the unit. A step in the inlet
temperature does not reach the outlet until the fluid has physically traversed the
exchanger, a pure delay of $t_d = L/u$ set by the flow path length over the velocity,
and only then does the first-order metal warm-up begin. The transfer function from
inlet to outlet temperature is therefore a delay in series with a lag, $G(s) =
e^{-t_d s}/(1 + \tau s)$, and the delay is what makes exchanger loops hard to tune: a
lag alone can be controlled tightly, but a delay adds phase lag that no controller can
undo, so the achievable loop bandwidth is capped near $1/t_d$ regardless of the metal
time constant. A long shell unit at low velocity can carry a transport delay of tens
of seconds, comparable to its metal lag, and a loop that ignores the delay and tunes
against the lag alone will oscillate. The design answers are to place the sensor close
to the outlet so the measured delay is the true process delay and not an added
measurement lag, to raise the velocity where the pressure-drop budget allows so $t_d$
falls, and to use a model-based or delay-compensating controller (a Smith predictor)
that knows the delay explicitly. The same delay-plus-lag structure recurs wherever a
process moves a fluid through a vessel, which is why exchanger dynamics are the worked
example the process-control chapter of Volume IX returns to.
\end{mastery}

\section{An end-to-end case study: a kerosene cooler}\pathtag{standard}

The methods of this chapter pay off when an engineer must turn a process
duty into a piece of equipment specified well enough to order. We take a
shell-and-tube kerosene cooler and carry it from stream data through the
duty, the log-mean difference, the overall coefficient built up from
films and wall and fouling, the required area, the tube count and shell
size, a pressure-drop check on both sides, and a verdict. The cooler is
a good vehicle because every move of the chapter appears in it once: the
duty balance, the LMTD with its correction factor, the series thermal
network, the area equation, the geometry that realises the area, and the
pumping check that decides whether the geometry is affordable to run.

\subsection{Duty and stream data}

A refinery side-stream of kerosene must be cooled before it goes to
storage. The hot stream is $\dot{m}_h = 10\,\text{kg/s}$ of kerosene
($c_{p,h} = 2.20\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$, density
$\rho_h = 800\,\text{kg/m}^3$, viscosity $\mu_h = 4.0\times 10^{-4}\,
\text{Pa}\cdot\text{s}$, conductivity $k_h = 0.13\,\si{\watt\per\meter
\per\kelvin}$), cooling from $T_{h,i} = 160\,\degC$ to $T_{h,o} = 60\,
\degC$. The cold stream is cooling-tower water entering at $T_{c,i} =
30\,\degC$, with the outlet to be held at or below $52\,\degC$ to limit
scaling. The duty is set by the hot-side balance,
\[
\dot{Q} = \dot{m}_h c_{p,h}(T_{h,i} - T_{h,o}) = 10 \times 2200 \times
100 = 2.20\,\text{MW}.
\]
The cooling-water flow follows from the cold-side balance at the
$52\,\degC$ outlet limit,
\[
\dot{m}_c = \frac{\dot{Q}}{c_{p,c}(T_{c,o} - T_{c,i})} = \frac{2.20
\times 10^6}{4180 \times (52 - 30)} = 23.9\,\text{kg/s}.
\]
We round the cooling-water flow to $24\,\text{kg/s}$, which trims the
outlet to $T_{c,o} = 30 + 2.20\times 10^6/(24\times 4180) = 51.9\,\degC$,
inside the limit.

\subsection{Log-mean difference and correction factor}

The exchanger is a one-shell-pass, two-tube-pass shell-and-tube unit,
the standard first choice for this service. The counter-flow terminal
differences are $\Delta T_1 = T_{h,i} - T_{c,o} = 160 - 51.9 = 108.1\,
\si{\kelvin}$ and $\Delta T_2 = T_{h,o} - T_{c,i} = 60 - 30 = 30\,
\si{\kelvin}$, so
\[
\Delta T_{\text{LMTD,CF}} = \frac{108.1 - 30}{\ln(108.1/30)} = \frac{
78.1}{1.281} = 61.0\,\si{\kelvin}.
\]
The correction-factor parameters are $P = (T_{c,o} - T_{c,i})/(T_{h,i} -
T_{c,i}) = 21.9/130 = 0.17$ and $R = (T_{h,i} - T_{h,o})/(T_{c,o} -
T_{c,i}) = 100/21.9 = 4.6$. For one shell pass and two tube passes these
values give $F \approx 0.94$, comfortably above the $0.75$ floor, so the
configuration reaches the terminal temperatures without resorting to two
shells in series. The effective driving difference is $F\,\Delta
T_{\text{LMTD,CF}} = 0.94 \times 61.0 = 57.3\,\si{\kelvin}$. Figure
\ref{fig:vol04ch06:case-study-temperature-profile} draws the two
stream profiles and the terminal differences for the counter-flow
reference.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-case-study-temperature-profile.tex}

\subsection{Overall coefficient from films, wall, and fouling}

The overall coefficient is assembled as a series thermal network
referred to the tube outer area. The kerosene runs in the tubes and the
water in the shell, the usual placement that keeps the dirtier
cooling water on the shell side where it is easier to clean. We take
tubes of outer diameter $d_o = 19\,\si{\milli\meter}$, wall thickness
$2\,\si{\milli\meter}$ (inner diameter $d_i = 15\,\si{\milli\meter}$),
carbon-steel wall conductivity $k_w = 50\,\si{\watt\per\meter\per
\kelvin}$. Estimated film coefficients for this service are a tube-side
(kerosene) value $h_i = 900\,\si{\watt\per\meter\squared\per\kelvin}$
and a shell-side (water) value $h_o = 5000\,\si{\watt\per\meter\squared
\per\kelvin}$. The fouling resistances, referred to their own surfaces,
are $R_{f,i} = 0.0004$ for the kerosene and $R_{f,o} = 0.0003\,\text{m}^2
\cdot\si{\kelvin}/\text{W}$ for the cooling water. Referred to the
outer area, the overall coefficient is
\[
\frac{1}{U_o} = \frac{1}{h_o} + R_{f,o} + \frac{d_o \ln(d_o/d_i)}{2 k_w}
+ \frac{d_o}{d_i}R_{f,i} + \frac{d_o}{d_i}\frac{1}{h_i}.
\]
Evaluating each term in $\text{m}^2\cdot\si{\kelvin}/\text{W}$: the
shell film is $1/5000 = 0.000200$; the shell fouling is $0.000300$; the
wall is $0.019\ln(19/15)/(2\times 50) = 0.019\times 0.237/100 =
0.0000450$; the tube fouling referred out is $(19/15)\times 0.0004 =
0.000507$; the tube film referred out is $(19/15)/900 = 0.001407$. The
sum is $0.002459\,\text{m}^2\cdot\si{\kelvin}/\text{W}$, so
\[
U_o = \frac{1}{0.002459} = 407\,\si{\watt\per\meter\squared\per\kelvin}.
\]
The tube-side kerosene film and the tube-side fouling together are
$0.001914$ of the $0.002459$ total, that is, $78\,\%$ of the resistance:
the kerosene side governs, and any move to improve this exchanger
(higher tube velocity, internal enhancement, more frequent cleaning)
should be spent there rather than on the already-good water side. Figure
\ref{fig:vol04ch06:resistance-stack} draws the five terms to scale, and the
picture makes the verdict immediate: the two kerosene-side terms fill most of the
bar and the water film is a sliver, so reading the stack rather than a single
tabulated coefficient is what points the improvement effort at the side that
actually governs.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-resistance-stack.tex}

\subsection{Required area and tube count}

The area follows from the design equation with the correction factor,
\[
A = \frac{\dot{Q}}{U_o F\,\Delta T_{\text{LMTD,CF}}} = \frac{2.20\times
10^6}{407 \times 57.3} = 94.3\,\text{m}^2.
\]
We add a $15\,\%$ margin against the film-coefficient estimates and
end-of-cycle fouling beyond the design allowance, giving a design area
of $A_d = 1.15 \times 94.3 = 108\,\text{m}^2$. The outer surface per
unit length of one tube is $\pi d_o = \pi \times 0.019 = 0.0597\,\text{
m}^2/\text{m}$. Choosing a tube length $L = 4.88\,\text{m}$ (a standard
$16\,\text{ft}$ tube), one tube provides $0.0597 \times 4.88 = 0.291\,
\text{m}^2$, so the tube count is
\[
N_t = \frac{A_d}{\pi d_o L} = \frac{108}{0.291} = 371 \to 372\,\text{
tubes}
\]
rounded up to an even number for the two-pass layout (186 tubes per
pass). On a triangular pitch of $p_t = 1.25\,d_o = 24\,\si{\milli\meter}$,
the bundle outer-tube-limit diameter for $372$ tubes is roughly
$D_{\text{otl}} = 0.637\,p_t\sqrt{N_t/\text{CTP}}$ with a two-pass
tube-count constant near $0.90$, which lands the shell inner diameter at
about $D_s = 0.58\,\text{m}$. Figure
\ref{fig:vol04ch06:case-study-tube-layout} shows the triangular layout,
the pitch, and the relation of the bundle to the shell. A $0.58\,\text{m}$
shell with $372$ tubes of $4.88\,\text{m}$ is a routine TEMA-standard
unit.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-case-study-tube-layout.tex}

\subsection{Pressure-drop check}

A design that meets the duty is still unbuilt until the pumping cost is
acceptable. On the tube side, the kerosene splits across $186$ tubes per
pass. The flow area per pass is $186 \times (\pi/4) d_i^2 = 186 \times
(\pi/4)(0.015)^2 = 0.0329\,\text{m}^2$, so the kerosene velocity is $u_t
= \dot{m}_h/(\rho_h A_{\text{pass}}) = 10/(800 \times 0.0329) = 0.38\,
\text{m/s}$. The tube Reynolds number is $Re = \rho_h u_t d_i/\mu_h =
800 \times 0.38 \times 0.015/(4.0\times 10^{-4}) = 1.14\times 10^4$,
turbulent, with a friction factor $f \approx 0.316\,Re^{-0.25} = 0.316
\times (1.14\times 10^4)^{-0.25} = 0.031$. The flow path is two passes of
$4.88\,\text{m}$, so $L_{\text{tube}} = 9.76\,\text{m}$, and the
straight-tube drop is
\[
\Delta p_t = f\frac{L_{\text{tube}}}{d_i}\frac{\rho_h u_t^2}{2} = 0.031
\times \frac{9.76}{0.015} \times \frac{800 \times 0.38^2}{2} = 0.031
\times 650 \times 57.8 = 1.17\,\text{kPa}.
\]
Adding return-header and nozzle losses (a customary allowance of about
four velocity heads for a two-pass unit, roughly $4 \times \rho_h
u_t^2/2 = 231\,\text{Pa}$) brings the tube side to about $1.4\,\text{
kPa}$, far inside a typical $50$ to $70\,\text{kPa}$ allowance. The
kerosene velocity is in fact low, which warns that the modest tube-side
film coefficient is the price of running well below the $1$ to $3\,\text{
m/s}$ liquid band; a designer chasing a smaller exchanger would raise the
tube-side velocity by using fewer, longer tubes, trading pressure drop
for area. The shell side, by the Kern method, sits in the same low range
for this flow, so neither side binds and the design is pumping-feasible.

\subsection{Verdict}

The kerosene cooler is a one-shell-pass, two-tube-pass shell-and-tube
unit of $372$ carbon-steel tubes, $19\,\si{\milli\meter}$ outer diameter,
$4.88\,\text{m}$ long, on a $24\,\si{\milli\meter}$ triangular pitch in
a shell of about $0.58\,\text{m}$ inner diameter, delivering the
$2.20\,\text{MW}$ duty at an overall coefficient near $407\,\si{\watt
\per\meter\squared\per\kelvin}$ with both pressure drops well inside
their allowances. The design is governed by the kerosene-side
resistance, carries a $15\,\%$ area margin for fouling and estimate
error, and would be confirmed against thermal-rating software before
purchase. The single number an engineer would sign is the area,
$108\,\text{m}^2$ installed, with the note that the cooler is
film-limited on the kerosene side and that the cooling-water outlet sits
at the scaling limit, so the cleaning interval is set by the water side
even though the heat-transfer bottleneck is the oil. Both observations
come straight out of the term-by-term thermal network, which is why the
network is worth assembling explicitly rather than reaching for a single
tabulated $U$.

\section{A second end-to-end case study: a milk pasteurizer with regeneration}\pathtag{standard}

The kerosene cooler sized a single shell to a single duty. A plate-and-frame
unit changes the problem in two ways worth carrying through end to end. The
exchanger is built from several sections on one frame, and the design's whole
economic argument turns on heat \emph{regeneration}: using the hot product
stream to preheat the cold feed stream so that external heating and cooling carry
only what regeneration cannot. We take a high-temperature short-time (HTST) milk
pasteurizer, the standard dairy unit, and carry it from the legal duty through
the regeneration ratio, the plate count of each section, the utility loads, and a
channel pressure-drop check. The pasteurizer is a good second vehicle because the
balanced-stream plate exchanger sits at the corner of the effectiveness-NTU map
where the close approach plate units are bought for becomes expensive, and the
regeneration ratio makes that cost visible as money.

\subsection{Duty, regeneration, and the legal constraint}

Raw milk enters at $T_{\text{raw}} = 4\,\degC$ and must be heated to the holding
temperature $T_{\text{hold}} = 72\,\degC$, held there for $15\,\text{s}$ (the
HTST legal minimum), then cooled back to $4\,\degC$ for packing. The product flow
is $\dot{m} = 5\,\text{kg/s}$ of milk ($c_p = 3.9\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$, close to water because milk is mostly water). The two streams
through the regeneration section are the same flow, so the capacity rates match
and the section runs balanced, $C_r = 1$, the case the close-approach plate unit
is designed for. The total heating load with no regeneration would be $\dot{Q}_
{\text{full}} = \dot{m} c_p (T_{\text{hold}} - T_{\text{raw}}) = 5\times 3900
\times (72 - 4) = 1.33\,\text{MW}$, and an equal cooling load would chill the
product back down, so a pasteurizer without regeneration would buy
$1.33\,\text{MW}$ of heating and $1.33\,\text{MW}$ of cooling for every batch.
The whole point of the regeneration section is to cut both. Figure
\ref{fig:vol04ch06:pasteurizer-flowsheet} shows the three-section arrangement:
the regeneration section trades heat between hot product and cold feed, a small
heating section lifts the feed the last few kelvin to $72\,\degC$, and a cooling
section takes the product the last stretch down to $4\,\degC$.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-pasteurizer-flowsheet.tex}

\subsection{Regeneration ratio and the plate count it demands}

The \engterm{regeneration ratio} is the fraction of the full sensible swing the
regeneration section recovers,
\[
R = \frac{T_{\text{preheat}} - T_{\text{raw}}}{T_{\text{hold}} - T_{\text{raw}}},
\]
the temperature the cold feed reaches in the regenerator over the temperature it
would reach if regeneration were complete. For a balanced ($C_r = 1$)
counter-current section the regeneration ratio is exactly the effectiveness,
$R = \varepsilon = \text{NTU}/(1 + \text{NTU})$, the relation plotted in figure
\ref{fig:vol04ch06:regeneration-ratio}. A modern dairy targets $R$ in the high
eighties to low nineties; we design for $R = 0.90$. The required transfer units
follow by inverting the balanced relation, $\text{NTU} = R/(1 - R) = 0.90/0.10 =
9.0$. The feed therefore leaves the regenerator preheated to $T_{\text{preheat}}
= T_{\text{raw}} + R(T_{\text{hold}} - T_{\text{raw}}) = 4 + 0.90\times 68 =
65.2\,\degC$, and the hot product, by the matched-flow symmetry, leaves the
regenerator cooled to $72 - 0.90\times 68 = 10.8\,\degC$.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-regeneration-ratio.tex}

The area follows from the transfer units. Plate units run high overall
coefficients because the corrugation drives turbulence at low velocity; take
$U = 3500\,\si{\watt\per\meter\squared\per\kelvin}$ for this clean milk-to-milk
service. The capacity rate of one stream is $C = \dot{m} c_p = 5\times 3900 =
19{,}500\,\si{\watt\per\kelvin}$, so the regeneration-section area is $A_{\text{
regen}} = \text{NTU}\,C/U = 9.0\times 19{,}500/3500 = 50.1\,\text{m}^2$. With a
plate projected area of $0.30\,\text{m}^2$ and heat transfer from both faces of
each interior plate, the area per thermal plate is about $0.30\,\text{m}^2$, so
the regeneration section needs roughly $A_{\text{regen}}/0.30 = 167$ thermal
plates, say $168$ to keep the port arrangement even. The steep cost of the close
approach shows here: lifting $R$ from $0.90$ to $0.95$ would raise NTU from $9$
to $19$ and the plate count from $168$ to about $355$, more than doubling the
regeneration stack to recover a further five percent. The dairy economics settle
in the high eighties to low nineties for exactly this reason.

\subsection{Utility loads after regeneration}

The heating section supplies only the gap the regenerator leaves. The feed
arrives at $65.2\,\degC$ and must reach $72\,\degC$, so the heating duty is
$\dot{Q}_{\text{heat}} = \dot{m} c_p (T_{\text{hold}} - T_{\text{preheat}}) =
5\times 3900\times (72 - 65.2) = 133\,\text{kW}$. The cooling section takes the
product from the $10.8\,\degC$ the regenerator left it at down to $4\,\degC$,
a duty of $\dot{Q}_{\text{cool}} = 5\times 3900\times (10.8 - 4) = 133\,\text{
kW}$. Regeneration has cut both utility loads from $1.33\,\text{MW}$ to
$133\,\text{kW}$, a tenfold reduction, which is the same factor as $1 - R =
0.10$. The regeneration ratio is therefore read twice: once as the effectiveness
that sets the plate count, and once as the fraction of utility the design still
buys. The heating section, sized for $133\,\text{kW}$ against hot water with a
modest approach, needs only a small fraction of the regeneration plate count;
the cooling section likewise. A pasteurizer is mostly regeneration plates by
count, which is why the unit's capital sits in the regeneration stack and its
operating cost sits in the small remainder.

\subsection{Channel pressure drop and the velocity the gaskets allow}

The plate channels are narrow, so pressure drop is the binding mechanical check.
Take a channel gap $b = 3\,\si{\milli\meter}$, a plate width $w = 0.5\,\text{m}$,
and the milk split across $N_c = 83$ parallel channels per pass on each side
(half of the $168$ plates form one side's channels). The flow area per side is
$A_{\text{ch}} = N_c\, b\, w = 83\times 0.003\times 0.5 = 0.125\,\text{m}^2$, so
the channel velocity is $u = \dot{m}/(\rho A_{\text{ch}}) = 5/(1020\times 0.125)
= 0.039\,\text{m/s}$ on a per-channel basis, low in absolute terms because the
many parallel channels share the flow. The hydraulic diameter of a thin slot is
$d_h \approx 2b = 6\,\si{\milli\meter}$, giving a channel Reynolds number $Re =
\rho u d_h/\mu = 1020\times 0.039\times 0.006/(1.5\times 10^{-3}) = 159$, in the
laminar-but-corrugation-enhanced range typical of plate units. Plate-channel
friction factors are correlated as $f \approx K Re^{-m}$ with chevron-dependent
constants; a representative $f \approx 18\,Re^{-0.3} = 18\times 159^{-0.3} = 3.9$
for a high-theta chevron captures the corrugation's large form drag. The
single-pass channel length is the plate flow length $L \approx 1.0\,\text{m}$, so
the per-pass drop is $\Delta p = f (L/d_h)\rho u^2/2 = 3.9\times (1.0/0.006)
\times 1020\times 0.039^2/2 = 3.9\times 167\times 0.776 = 505\,\text{Pa}$ per
pass. The regeneration section uses several passes in series to reach
$\text{NTU} = 9$; at, say, four passes per side the regeneration pressure drop is
about $2\,\text{kPa}$, and the whole frame including ports and the heating and
cooling sections lands near $10$ to $15\,\text{kPa}$ per stream, inside the
gasketed unit's customary allowance. The check that governs is not the pressure
drop in absolute terms but the channel velocity against the gasket pressure
rating ($\sim 25\,\text{bar}$ for nitrile) and against the velocity that keeps
the milk-side film high enough to hold the assumed $U$; too few channels raises
the velocity and the film but risks the pressure drop, too many drops the film
and forces more plates, the same area-versus-pumping trade the chapter keeps
returning to, now read on the channel rather than the shell.

\subsection{Verdict}

The pasteurizer is a single gasketed plate-and-frame frame carrying about $168$
regeneration plates plus smaller heating and cooling sections, running a balanced
regeneration section at $R = 0.90$ with an overall coefficient near $3500\,
\si{\watt\per\meter\squared\per\kelvin}$, delivering the legal $72\,\degC$ hold
on $5\,\text{kg/s}$ of milk while buying only $133\,\text{kW}$ of heating and
$133\,\text{kW}$ of cooling against the $1.33\,\text{MW}$ each that no
regeneration would demand. The single number an engineer would defend is the
regeneration ratio, because it sets the plate count, the utility bill, and the
sensitivity of both: a small move in $R$ moves the plate count steeply and the
utility load gently, so the design is read off the flat upper part of the
effectiveness-NTU curve where a further point of regeneration is no longer worth
its plates. The contrast with the kerosene cooler is the lesson: the shell unit
was sized to a duty and checked for pumping, while the plate unit is sized to a
\emph{recovery target} whose economics, not whose duty, fix the area.

\section{A third end-to-end case study: an air-cooled process gas cooler}\pathtag{standard}

The shell-and-tube unit rejected its heat to cooling water and the plate unit to a
utility loop; both assumed a cold-side liquid waiting to carry the heat away. An
air-cooled cooler rejects heat straight to the atmosphere, which removes the
water-supply and water-treatment burden but replaces a high liquid film with a
low air film and a fan that must be fed power for the life of the plant. We size a
forced-draft fin-fan cooler for a process gas duty and carry it from the duty
through the air-side finned-surface coefficient, the overall coefficient on the
finned area, the required area and tube count, the face area the fan must serve,
and the fan power that is the unit's running cost. The air cooler is the right
third vehicle because it exercises the fin algebra of the preceding section, it
moves the binding resistance onto a gas side where extended surface earns its
keep, and it makes the operating cost a first-class design variable rather than an
afterthought. Figure \ref{fig:vol04ch06:aircooler-arrangement} shows the
arrangement.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-aircooler-arrangement.tex}

\subsection{Duty and the air the fan must move}

A gas-processing plant in a hot, dry location must cool $\dot{m}_h = 9\,\text{kg/s}$
of process gas ($c_{p,h} = 2.4\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$) from
$T_{h,i} = 120\,\degC$ to $T_{h,o} = 55\,\degC$ before the next stage. The duty is
\[
\dot{Q} = \dot{m}_h c_{p,h}(T_{h,i} - T_{h,o}) = 9\times 2400\times 65 = 1.40\,
\text{MW}.
\]
The cold side is ambient air, and the design ambient for the site is a hot
$T_{c,i} = 40\,\degC$, the worst-case temperature the cooler must still meet, the
air-cooled counterpart of the cooling tower's design wet-bulb. We move
$\dot{m}_{\text{air}} = 70\,\text{kg/s}$ of air, which warms by $\Delta T_{\text{
air}} = \dot{Q}/(\dot{m}_{\text{air}} c_{p,\text{air}}) = 1.40\times 10^6/(70\times
1005) = 20.0\,\si{\kelvin}$, leaving the cooler at $60\,\degC$. The capacity rates
are $C_h = 9\times 2400 = 21{,}600\,\si{\watt\per\kelvin}$ for the gas and
$C_{\text{air}} = 70\times 1005 = 70{,}350\,\si{\watt\per\kelvin}$ for the air, so
$C_{\min} = 21{,}600$ (the gas), $C_{\max} = 70{,}350$, and $C_r = 0.31$. The gas
governs the maximum possible duty, $\dot{Q}_{\max} = C_{\min}(T_{h,i} - T_{c,i}) =
21{,}600\times 80 = 1.73\,\text{MW}$, so the required effectiveness is
$\varepsilon = \dot{Q}/\dot{Q}_{\max} = 1.40/1.73 = 0.81$. An effectiveness of
$0.81$ at $C_r = 0.31$ is a demanding but reachable single-unit target, which the
NTU will confirm.

\subsection{Air-side finned-surface coefficient}

The air film is the cooler's weak link, so the tubes carry fins to multiply the
air-side area. We take an air-side film coefficient on the finned surface of
$h_o = 70\,\si{\watt\per\meter\squared\per\kelvin}$, an aluminium fin field whose
fin area is $A_f/A = 0.92$ of the total air-side area, and a fin efficiency
$\eta_f = 0.85$ for the chosen fin height and the aluminium metal (the single-fin
worked example showed how that $0.85$ comes from the group $mL$). The overall
surface efficiency is
\[
\eta_o = 1 - \frac{A_f}{A}(1 - \eta_f) = 1 - 0.92\times(1 - 0.85) = 0.862,
\]
so the effective air-side conductance is $\eta_o h_o = 0.862\times 70 = 60.3\,
\si{\watt\per\meter\squared\per\kelvin}$, the bare $70$ docked by the
fourteen-percent cost of conducting heat out along the fins. Treating the whole
finned field as fully effective would have credited the design with $70$ instead
of $60$, a silent over-statement of the dominant air-side conductance, which is
the error the surface-efficiency factor exists to prevent.

\subsection{Overall coefficient on the finned area}

The network is referred to the total air-side (finned) area, the basis on which
air-cooler areas are quoted. The fins multiply the air-side area to about
$A_o/A_i = 16$ times the inside bare tube area, so every tube-side term is
amplified by that ratio when referred out. The process-gas film inside the tubes
is $h_i = 700\,\si{\watt\per\meter\squared\per\kelvin}$ for the compressed gas at
this velocity, the tube-side fouling is $R_{f,i} = 0.0002\,\text{m}^2\cdot\si{
\kelvin}/\text{W}$, and the carbon-steel tube ($d_o = 25.4\,\si{\milli\meter}$,
$d_i = 21.2\,\si{\milli\meter}$, $k_w = 50\,\si{\watt\per\meter\per\kelvin}$)
contributes a small wall term. Evaluating each contribution in
$\text{m}^2\cdot\si{\kelvin}/\text{W}$:
\[
\frac{1}{U_o} = \underbrace{\frac{1}{\eta_o h_o}}_{0.0166}
+ \underbrace{\frac{A_o}{A_i}\frac{d_i\ln(d_o/d_i)}{2k_w}}_{0.0006}
+ \underbrace{\frac{A_o}{A_i}R_{f,i}}_{0.0032}
+ \underbrace{\frac{A_o}{A_i}\frac{1}{h_i}}_{0.0229}.
\]
The sum is $0.0433\,\text{m}^2\cdot\si{\kelvin}/\text{W}$, so
\[
U_o = \frac{1}{0.0433} = 23.1\,\si{\watt\per\meter\squared\per\kelvin}
\]
on the finned area, the low coefficient that air coolers always show because the
area they are referred to is so large. The two governing resistances are the
air-side film ($0.0166$, about $38\,\%$) and the tube-side film referred out
($0.0229$, about $53\,\%$); the two are comparable, which is the balanced
condition good air-cooler practice aims for, because spending more fin to lift the
air side past the tube side would buy area on the side that no longer governs. If
the tube-side film were doubled by a higher gas velocity, the air side would
become the sole bottleneck and the fin field would have to grow; if the fin field
were trimmed, the air side would govern and the gas-side velocity would be wasted.

\subsection{Required area, tube count, and face area}

The effectiveness target sets the transfer units. Inverting the cross-flow
both-unmixed relation at $\varepsilon = 0.81$ and $C_r = 0.31$ gives
$\text{NTU} \approx 2.19$, so the required conductance is $U_o A_o = \text{NTU}\,
C_{\min} = 2.19\times 21{,}600 = 47{,}300\,\text{W}/\si{\kelvin}$ and the finned
area is
\[
A_o = \frac{\text{NTU}\,C_{\min}}{U_o} = \frac{47{,}300}{23.1} = 2050\,\text{m}^2.
\]
At an area ratio of $16$, the bare tube area is $A_o/16 = 128\,\text{m}^2$, which
at $\pi d_o = \pi\times 0.0254 = 0.0798\,\text{m}^2/\text{m}$ of bare outer surface
per metre is $128/0.0798 = 1600\,\text{m}$ of tube. Choosing a $10\,\text{m}$ tube
length gives $1600/10 = 160$ tubes, a routine air-cooler bundle of perhaps four to
six rows deep on a triangular pitch. The face area the air must pass is set by the
face mass velocity: at an air density near $1.09\,\text{kg/m}^3$ at the mean air
temperature and a customary face velocity of $3\,\text{m/s}$, the face area is
\[
A_{\text{face}} = \frac{\dot{m}_{\text{air}}}{\rho_{\text{air}} u_{\text{face}}}
= \frac{70}{1.09\times 3.0} = 21.4\,\text{m}^2,
\]
about $4\,\text{m}$ by $5\,\text{m}$, served by two fans for redundancy and
turndown. Figure \ref{fig:vol04ch06:aircooler-rating} plots the duty and the gas
outlet temperature against the air flow the fans deliver and marks this design
point.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-aircooler-rating.tex}

\subsection{Fan power, the running cost}

The fan power is the price of moving the air, and it grows as the cube of the air
flow, so it is the variable that punishes over-cooling. The air pressure drop
across a finned bundle of this depth is of order $150\,\text{Pa}$. The volumetric
air flow is $\dot{V}_{\text{air}} = \dot{m}_{\text{air}}/\rho_{\text{air}} =
70/1.09 = 64\,\text{m}^3/\text{s}$, so the air-side hydraulic power is
$\dot{V}_{\text{air}}\,\Delta p = 64\times 150 = 9.6\,\text{kW}$, and at a combined
fan-and-motor efficiency of about $0.65$ the shaft power is
\[
\dot{W}_{\text{fan}} = \frac{\dot{V}_{\text{air}}\,\Delta p}{\eta} =
\frac{9.6}{0.65} = 14.8\,\text{kW}.
\]
The $15\,\text{kW}$ of fan power runs whenever the cooler runs, so over an
$8000\,\text{hour}$ year it draws about $120\,\text{MWh}$, the operating cost the
design trades against the capital of the finned bundle. The rating curve shows why
the design does not simply move more air: pushing the air flow from $70$ to
$100\,\text{kg/s}$ would lower the gas outlet by only a few kelvin while raising
the fan power by the cube of the flow ratio, $(100/70)^3 = 2.9$ times, to over
$40\,\text{kW}$. The marginal degree of cooling stops earning its fan power near
the chosen point, which is where the design sits.

\subsection{Verdict}

The cooler is a forced-draft air-cooled unit of about $160$ carbon-steel finned
tubes, $25.4\,\si{\milli\meter}$ outer diameter, $10\,\text{m}$ long, with an
aluminium fin field carrying $2050\,\text{m}^2$ of finned area on a
$21\,\text{m}^2$ face, served by two fans drawing about $15\,\text{kW}$ at the
design point and delivering the $1.40\,\text{MW}$ duty against a $40\,\degC$ design
ambient. The single number an engineer would defend is the design ambient
temperature, because it fixes the approach the cooler must make and therefore the
area: a cooler sized for a $40\,\degC$ ambient still meets its gas outlet on the
few summer hours that touch it, while one sized for an annual-average ambient would
fail when the plant most needs cooling. The contrast with the two earlier cases is
the cold side: the shell unit and the plate unit drew on a cold-side fluid whose
temperature the plant controlled, while the air cooler is at the mercy of the
weather, so its design point is a worst-case ambient and its running cost is fan
power rather than pumping a liquid. The fin algebra of the preceding section is
what made the air-side area credible, and the balanced air-side and tube-side
resistances are the sign that the fin field is neither starved nor wasted.

\section{An applied vignette: the automotive radiator}\pathtag{standard}

The three case studies sized industrial units against design points an operator
sets. A car radiator runs the same physics against a cold-side flow the driver
sets with the accelerator, and it is worth a short pass because it shows the
effectiveness method on a unit everyone has met and because its sizing point is a
worst case rather than a steady duty. A radiator is a compact cross-flow exchanger:
hot coolant (a water-glycol mix) runs through flattened tubes, and ram air plus
fan-driven air crosses the finned core. We size it for the duty an engine rejects
to coolant, then read why the binding case is the traffic jam, not the highway.

\subsection{Duty and the air the core must pass}

A mid-size engine producing $80\,\text{kW}$ at the wheels rejects roughly a third
of its fuel energy to the coolant, so the radiator duty at a sustained climb is of
order $\dot{Q} = 60\,\text{kW}$. The coolant flow is $\dot{m}_c = 1.5\,\text{kg/s}$
of water-glycol ($c_p = 3.6\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$), entering the
radiator from the engine at $T_{h,i} = 95\,\degC$ and, with the thermostat full
open, leaving cooler to return to the block. The capacity rate of the coolant is
$C_c = 1.5\times 3600 = 5400\,\si{\watt\per\kelvin}$. On a $35\,\degC$ summer day
the air enters at $T_{c,i} = 35\,\degC$, the design ambient that plays the role the
gas cooler's $40\,\degC$ and the cooling tower's wet-bulb played. At highway speed
the core passes a large ram-air flow, say $\dot{m}_{\text{air}} = 2.4\,\text{kg/s}$,
so $C_{\text{air}} = 2.4\times 1005 = 2412\,\si{\watt\per\kelvin}$; the air is the
smaller capacity rate, $C_{\min} = 2412$, and $C_r = 2412/5400 = 0.45$. The maximum
possible duty is $\dot{Q}_{\max} = C_{\min}(T_{h,i} - T_{c,i}) = 2412\times 60 =
145\,\text{kW}$, so the required effectiveness at the $60\,\text{kW}$ duty is
$\varepsilon = 60/145 = 0.41$, an easy target at highway speed where ram air is
plentiful.

\subsection{The worst case is the idle, not the cruise}

The radiator is not sized for the cruise. At a standstill in slow traffic the road
gives no ram air, and the engine-driven or electric fan alone must move the air. A
fan that moves $\dot{m}_{\text{air}} = 0.7\,\text{kg/s}$ gives $C_{\text{air}} =
0.7\times 1005 = 704\,\si{\watt\per\kelvin}$, now far below the coolant, so
$C_{\min} = 704$ and the maximum possible duty falls to $\dot{Q}_{\max} = 704\times
60 = 42\,\text{kW}$, which is below the $60\,\text{kW}$ the engine still rejects if
it is working (a long idle with the air-conditioning compressor loaded, or creeping
up a grade in traffic). The radiator that coasted at highway speed is now starved of
air, the coolant cannot shed its heat, and the engine temperature climbs toward
boil-over. The figure \ref{fig:vol04ch06:radiator-rating} plots the duty against
road speed and marks the two points: the comfortable highway cruise and the
fan-only idle that sets the size. The design must reach the rejected duty at the
idle point, which fixes the fan capacity and the core face area, and the highway
case then has ample margin. The lesson repeats the gas cooler's: an air-side
exchanger is sized against the worst-case cold-side flow and ambient, and for a
radiator that worst case is the one moment the car is not moving.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-radiator-rating.tex}

\begin{mastery}
The radiator shares its air with the cabin heater and the air-conditioning
condenser, and the three form a small thermal system whose interactions set real
design constraints. The cabin heater is a second small exchanger fed from the same
coolant loop, so it is a deliberate parasitic load that helps the radiator on a
cold day and is shut off by a valve on a hot one. The air-conditioning condenser
sits in front of the radiator in the air path, so it preheats the cooling air
before that air reaches the radiator, which is the reason the worst thermal case
for an engine is a hot day in traffic with the air-conditioning running: the
condenser is dumping its heat into the very air the radiator needs cold, and the
fan is the only air mover. Modern vehicles manage this with a controlled electric
fan, a coolant thermostat that is itself a temperature-actuated valve, and on
many cars a grille shutter that closes at speed to cut aerodynamic drag once ram
air is no longer needed, trading a small cooling margin for fuel economy. The
radiator is therefore not a standalone exchanger but a node in a controlled thermal
network, which is the same step up from single unit to system that the district
network and the pinch construction make at industrial scale.
\end{mastery}

\section{An applied vignette: the exhaust-gas recirculation cooler}\pathtag{standard}

The radiator rejected engine heat to keep the block cool; a second small exchanger
on the same engine cools a slip stream of the engine's own exhaust for a different
reason, and it is worth a short pass because it shows an exchanger sized by a
downstream chemistry constraint rather than by a duty an operator picks. An
exhaust-gas recirculation (EGR) cooler takes a fraction of the hot exhaust,
cools it against engine coolant, and returns it to the intake, where the cooled
inert gas lowers the peak combustion temperature and cuts the nitrogen-oxide
formation that runs away above a threshold flame temperature. The cooler's target
outlet temperature is set by that chemistry, not by a heat balance, which turns the
sizing around.

\subsection{Duty against a chemistry target}

A diesel engine recirculates $\dot{m}_g = 0.06\,\text{kg/s}$ of exhaust gas
($c_{p,g} = 1.10\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$) taken off at
$T_{g,i} = 550\,\degC$, and the intake charge target asks that it return at
$T_{g,o} = 150\,\degC$. The duty follows from the gas-side balance,
\[
\dot{Q} = \dot{m}_g c_{p,g}(T_{g,i} - T_{g,o}) = 0.06\times 1100\times 400 =
26.4\,\text{kW}.
\]
The cold side is engine coolant at $T_{c,i} = 90\,\degC$ (the block runs hot),
flowing at $\dot{m}_c = 0.25\,\text{kg/s}$ of water-glycol
($c_p = 3.7\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$), so the coolant capacity rate
is $C_c = 0.25\times 3700 = 925\,\si{\watt\per\kelvin}$ against a gas capacity rate
of $C_g = 0.06\times 1100 = 66\,\si{\watt\per\kelvin}$. The gas is by far the
smaller, so $C_{\min} = 66$, $C_r = 66/925 = 0.071$, nearly the phase-change limit,
and the coolant warms only by $\dot{Q}/C_c = 26{,}400/925 = 28.5\,\si{\kelvin}$ to
leave at $118.5\,\degC$, staying liquid under the cooling-system pressure. The
maximum possible duty is $\dot{Q}_{\max} = C_{\min}(T_{g,i} - T_{c,i}) = 66\times
(550 - 90) = 30.4\,\text{kW}$, so the required effectiveness is $\varepsilon =
26.4/30.4 = 0.87$, a demanding target that the near-zero $C_r$ makes reachable, since
at $C_r \to 0$ the effectiveness climbs as $1 - e^{-\text{NTU}}$ with no ceiling
below one.

\subsection{The area and the fouling that governs its life}

At $C_r \approx 0$ the transfer units follow from $\varepsilon = 1 -
e^{-\text{NTU}}$, so $\text{NTU} = -\ln(1 - 0.87) = -\ln(0.13) = 2.04$, and the
conductance is $UA = \text{NTU}\,C_{\min} = 2.04\times 66 = 135\,\text{W}/\si{
\kelvin}$. The gas-side film is the weak link, of order $h_g = 120\,\si{\watt\per
\meter\squared\per\kelvin}$ for the recirculated gas at its velocity, against a
liquid coolant film an order of magnitude higher, so the coefficient on the gas
area is near $U = 110\,\si{\watt\per\meter\squared\per\kelvin}$ and the gas-side
area is $A = 135/110 = 1.2\,\text{m}^2$, realised as a small stainless tube bundle
or a stack of stamped plates that fits in a fist-sized housing. The governing life
constraint is soot and hydrocarbon fouling on the gas side: the cooled exhaust
lays down a carbon deposit that, unlike the asymptotic waterside scale of an
industrial unit, keeps growing and can cut the duty enough that the intake charge
returns too hot and the nitrogen-oxide control drifts out of specification. The
design answer is a gas-side geometry that runs the gas fast enough to scour the
deposit, a stainless surface the acidic condensate does not corrode, and a duty
margin that tolerates an end-of-service-interval deposit, the same over-design logic
the failure section applies to industrial fouling, here driven by an emissions
limit rather than a throughput loss.

\subsection{Why the sizing is inverted}

The EGR cooler is sized backward from the three earlier cases. The kerosene cooler
and the air cooler took a required duty and found the area; the EGR cooler takes a
required gas outlet temperature, fixed by combustion chemistry downstream, and the
duty and area fall out of it. The coolant is not chosen for its temperature but is
whatever the engine's cooling loop already carries, so the cooler must make its
approach against a $90\,\degC$ cold side that a process engineer would consider
uncomfortably warm, which is exactly why the near-zero $C_r$ and the high
effectiveness are forced rather than chosen. The unit closes the chapter's arc from
the process plant to the vehicle: the same LMTD and effectiveness relations size a
$100\,\text{m}^2$ refinery shell and a $1\,\text{m}^2$ under-hood bundle, and only
the binding constraint, cost and pumping in one case, chemistry and fouling in the
other, changes what the engineer signs.

\section{An applied vignette: the rotary regenerator}\pathtag{standard}

Every exchanger so far passes heat through a wall that separates the two streams
at all times. A regenerator breaks that rule: it stores heat in a solid matrix
that the two streams sweep in turn, so the same metal that took heat from the hot
stream a moment ago gives it back to the cold stream a moment later. The rotary
regenerator, or thermal wheel, is the common realisation, and it is worth a short
pass because it sizes on a stored-heat balance rather than a steady wall
conductance and because its one weakness, carry-over between the streams, decides
where it may be used. Figure \ref{fig:vol04ch06:rotary-regenerator} shows the
arrangement: a porous matrix wheel turns slowly, one half swept by the hot exhaust
and the other by the cold supply, each matrix element charging on the hot side and
discharging on the cold side once per revolution.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-rotary-regenerator.tex}

\subsection{Duty and the effectiveness a wheel reaches}

A building ventilation system exhausts $\dot{m} = 3\,\text{kg/s}$ of warm stale
air at $T_{h,i} = 22\,\degC$ and draws in the same mass flow of cold fresh air at
$T_{c,i} = -8\,\degC$ on a winter day, and the wheel recovers heat from the
exhaust to preheat the intake. The two streams are matched air flows, so the
capacity rates are equal, $C_h = C_c = 3\times 1005 = 3015\,\si{\watt\per\kelvin}$,
and the wheel runs balanced at $C_r = 1$, the same corner the dairy regenerator
lived in. A well-designed thermal wheel reaches an effectiveness of about
$\varepsilon = 0.80$ at a matrix mass and rotation speed that keep the matrix
heat capacity large against the per-revolution heat swing. The recovered duty is
$\dot{Q} = \varepsilon C_{\min}(T_{h,i} - T_{c,i}) = 0.80\times 3015\times
(22 - (-8)) = 0.80\times 3015\times 30 = 72\,\text{kW}$, so the intake air is
preheated by $\varepsilon(T_{h,i} - T_{c,i}) = 0.80\times 30 = 24\,\si{\kelvin}$
to leave the wheel at $-8 + 24 = 16\,\degC$, a rise that cuts the building's
heating coil load to the small remainder between $16\,\degC$ and the supply
setpoint. The $72\,\text{kW}$ recovered is heat the boiler no longer burns fuel to
supply, the same recovery economics the economiser and the pasteurizer showed, now
carried by a rotating solid rather than a fixed wall.

\subsection{The matrix mass and the rotation speed}

The wheel works only if each matrix element stores enough heat to carry a useful
swing across the seal without changing its own temperature much. The heat a matrix
element exchanges per revolution is the duty divided by the rotation rate, so a
faster wheel moves smaller parcels more often and holds a flatter matrix
temperature, which raises the effectiveness toward the counter-flow ceiling. The
matrix heat capacity rate, the product of the matrix mass, its specific heat, and
the rotation rate, must be large against the stream capacity rates for the wheel to
behave like a true counter-flow exchanger; the governing dimensionless group is the
matrix capacity-rate ratio $C_r^\ast = (m_{\text{matrix}} c_{\text{matrix}}
N)/C_{\min}$, and effectiveness climbs with $C_r^\ast$ toward a ceiling set by the
transfer units and then flattens. Design practice sets the rotation near ten to
twenty revolutions per minute, fast enough that $C_r^\ast$ exceeds about five and
the effectiveness sits within a point or two of its ceiling, slow enough that the
carry-over volume swept across per revolution stays small. Push the wheel slower
and the matrix swings too far in temperature, the hot end of each element cooling
below the cold stream it is meant to warm, and the effectiveness falls; push it
faster and the seal wear and the carry-over both climb with no effectiveness left
to gain.

\subsection{Carry-over, the constraint that places the wheel}

The rotary regenerator's one structural weakness is that the matrix pores trap a
sliver of whichever stream last swept them and drag it across the seal into the
other stream. For the ventilation wheel this carry-over is a small fraction of a
percent of the flow, harmless when the two streams are both building air, so the
thermal wheel is the standard heat-recovery device for balanced ventilation. The
same carry-over rules the wheel out wherever the two streams must never mix: a
process exhaust carrying a contaminant, a laboratory fume stream, or a supply that
feeds a clean room cannot tolerate even a fraction of a percent of the other
stream, and there a fixed-plate recuperator or a run-around coil loop replaces the
wheel at the cost of a lower effectiveness. A purge sector, a small wedge of the
wheel swept by clean air between the hot and cold zones, cuts the carry-over by an
order of magnitude and extends the wheel to more sensitive services, at the cost
of the purge air the fan must move. The vignette closes the same way the EGR cooler
did: the heat-transfer relation is routine, and the binding constraint, here
carry-over rather than fouling or chemistry, is what decides whether the device
may be used at all.

\begin{mastery}
The rotary regenerator's effectiveness formula is not the recuperator's, because a
regenerator adds a second capacity ratio the recuperator has no analogue for. A
recuperator is fixed by $\text{NTU}$ and the stream ratio $C_r$; a regenerator is
fixed by those two and by the matrix capacity-rate ratio $C_r^\ast$, and the
standard correlation writes the regenerator effectiveness as the counter-flow
recuperator value at the same $\text{NTU}$ and $C_r$, multiplied by a correction
$1 - 1/(9\,{C_r^\ast}^{1.93})$ that approaches unity as the matrix is made heavier
or turned faster. The correction is the quantitative form of the earlier statement
that a slow, light wheel behaves worse than a fast, heavy one: at $C_r^\ast = 1$
the correction docks the effectiveness by about eleven percent, while at
$C_r^\ast = 5$ it costs under half a percent, which is why design lives above
$C_r^\ast \approx 5$. The fixed-matrix regenerator, the checkerwork brick stove
that a blast furnace or a glass tank fires in alternating cycles, obeys the same
storage physics with the roles of rotation and valve-switching exchanged: instead
of a wheel carrying the matrix between two steady streams, a pair of fixed matrices
is switched between the hot and cold streams on a timed cycle, and the cycle period
plays the part the rotation rate plays in the wheel. Both are regenerators, both
size on a stored-heat balance, and both trade a carry-over or a switching
transient for the ability to run at temperatures and dirtiness that would foul or
melt a thin recuperator wall.
\end{mastery}

\section{An applied vignette: the heat pipe}\pathtag{standard}

The exchangers to this point move heat by forcing a fluid through a device with a
pump or a fan. A heat pipe moves heat with no moving part at all, by evaporating a
working fluid at the hot end and condensing it at the cold end, and it is worth a
short pass because it shows a device whose effective conductance dwarfs solid metal
and whose limits are set by capillary action rather than by a film coefficient.
Figure \ref{fig:vol04ch06:heat-pipe} shows the arrangement: an evaporator end
where heat boils the fluid, an adiabatic section where vapour flows to the cold
end, a condenser end where the vapour gives up its latent heat, and a wick that
returns the condensate by capillary suction.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-heat-pipe.tex}

\subsection{The effective conductivity a heat pipe reaches}

A copper rod carries heat by conduction, and its thermal resistance for a length
$L$ and cross-section $A$ is $L/(k A)$ with $k \approx 400\,\si{\watt\per\meter
\per\kelvin}$. A heat pipe carries the same heat by latent transport, and its
temperature drop end to end is only the small difference needed to drive the
vapour flow and the two phase changes. Take a heat pipe of length $L = 0.3\,\text{m}$
and cross-section $A = 1\,\text{cm}^2 = 10^{-4}\,\text{m}^2$ carrying
$\dot{Q} = 50\,\text{W}$ at an end-to-end temperature drop of $\Delta T = 2\,\si{
\kelvin}$. Its effective conductance is $\dot{Q}/\Delta T = 50/2 = 25\,\text{W}/
\si{\kelvin}$, so the effective conductivity that would give this conductance in a
solid rod of the same geometry is $k_{\text{eff}} = (\dot{Q}/\Delta T)(L/A) =
25\times 0.3/10^{-4} = 7.5\times 10^4\,\si{\watt\per\meter\per\kelvin}$, about
$190$ times copper. The heat pipe is not conducting; it is carrying latent heat in
a vapour and returning the condensate in a wick, and the enormous apparent
conductivity is the mark of a latent-transport device masquerading as a
conductor. A solid copper rod of the same size carrying $50\,\text{W}$ would drop
$\Delta T = \dot{Q} L/(k A) = 50\times 0.3/(400\times 10^{-4}) = 375\,\si{\kelvin}$
end to end, which is why electronics thermal management reaches for heat pipes to
spread a processor's heat to a finned sink rather than trying to conduct it through
a metal bar.

\subsection{The capillary limit that caps the heat carried}

A heat pipe cannot carry unlimited heat, because the wick must pump the condensate
back against the pressure the vapour flow and gravity impose, and the capillary
suction the wick pores generate is finite. The capillary pressure a pore of
effective radius $r_c$ can raise is $\Delta p_{\text{cap}} = 2\sigma/r_c$, with
$\sigma$ the surface tension of the working fluid, and this must exceed the sum of
the vapour-flow pressure drop, the liquid-flow pressure drop in the wick, and the
gravitational head if the condenser sits below the evaporator. Take water as the
working fluid near $60\,\degC$, where $\sigma \approx 0.066\,\text{N/m}$, in a
wick of pore radius $r_c = 50\,\si{\micro\meter}$: the capillary pressure is
$\Delta p_{\text{cap}} = 2\times 0.066/(50\times 10^{-6}) = 2640\,\text{Pa}$. If
the heat pipe must lift condensate against a gravitational head of even $0.1\,\text{m}$,
the head is $\rho g h = 983\times 9.81\times 0.1 = 964\,\text{Pa}$, so more than a
third of the capillary budget is spent on gravity alone, leaving the rest for the
vapour and liquid flows that scale with the heat carried. Push the heat rate up and
the flow pressure drops rise until they exhaust the capillary budget; at that point
the wick can no longer return all the condensate, the evaporator dries out, and the
pipe fails to carry more heat. This capillary limit is the reason a heat pipe is
rated for a maximum heat rate at a maximum adverse tilt, and it is why the working
fluid is matched to the temperature range: water for moderate electronics
temperatures, ammonia or methanol for cooler service, sodium or potassium for the
high temperatures of a re-entry leading edge or a nuclear application.

\subsection{Why the device is passive and where that matters}

The heat pipe closes the chapter's device survey with the one exchanger that needs
no external power to move its working fluid. The wick pumps the condensate, the
latent heat carries the load, and the only energy the device consumes is the small
temperature drop it costs, so a heat pipe runs as long as there is a temperature
difference to drive it and fails only when the capillary or boiling limit is
reached. That passivity is what places it: a spacecraft radiator that cannot carry
a pump, a permafrost thermosyphon that keeps a pipeline foundation frozen with no
power in a remote arctic site, and a laptop that cannot spare the space or the
noise of a liquid loop all reach for the heat pipe precisely because it asks for
nothing but a temperature difference. The contrast with the pumped and fanned
exchangers of the earlier sections is the lesson: those devices buy a high
coefficient with pumping power that runs for the life of the plant, while the heat
pipe buys an enormous effective conductance for free and pays instead with a hard
ceiling on the heat it can carry.

\section{An applied vignette: the spiral heat exchanger}\pathtag{standard}

The shell-and-tube unit splits one stream across a bundle of parallel tubes, and a
bundle is exactly what a fouling slurry blocks selectively, plugging some tubes while
others run clear until the maldistribution wrecks the duty. A spiral heat exchanger
takes the opposite construction, and it is worth a short pass because it shows a
geometry chosen for the fluids that defeat every unit so far: viscous, fibrous, and
solids-laden streams. Figure \ref{fig:vol04ch06:spiral-exchanger} shows the
arrangement, two metal sheets wound into a pair of interleaved spiral channels so
that each stream has a single long curved passage rather than a bundle of parallel
ones.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-spiral-exchanger.tex}

\subsection{The single passage that will not plug}

A spiral exchanger runs each stream through one continuous channel that spirals from
the periphery to the core, so there is no parallel path for the flow to abandon: if a
patch of the channel begins to foul, the whole stream must still pass it, and the
local velocity rises as the passage narrows, which scours the deposit back off. The
self-scouring action is the unit's defining virtue. Take a channel of gap
$b = 12\,\si{\milli\meter}$ and width $w = 0.6\,\text{m}$ carrying a viscous
$8\,\text{kg/s}$ sludge of density $\rho = 1050\,\text{kg/m}^3$; the channel velocity
is $u = \dot{m}/(\rho b w) = 8/(1050\times 0.012\times 0.6) = 1.06\,\text{m/s}$, held
high by the single narrow passage where a shell-and-tube unit spreading the same flow
across a wide bundle would drop to a settling velocity. The continuous curvature adds
a secondary swirl (a Dean-vortex circulation set up by the channel's curvature) that
lifts the film coefficient above a straight channel of the same gap and, more to the
point, keeps particles suspended rather than letting them settle on the wall. The
result is a unit that holds its duty on a service that would foul a bundle solid, at
the cost of a wide channel gap that lowers the area density well below a plate unit's.

\subsection{Where the geometry binds}

The spiral's price is the mirror of its virtue. The wound sheets fix the geometry once
and for all: the unit cannot be re-bundled, re-tubed, or re-gasketed the way a
shell-and-tube or plate unit can, so a spiral that outgrows its duty or corrodes
through is replaced whole. The pressure envelope is set by the wound sheets and their
welds rather than by a code-stamped shell, which caps the spiral well below the
pressure and temperature a shell-and-tube unit reaches, so the spiral is a
moderate-condition device. Cleaning, when the self-scouring is not enough, means
opening the covers and washing the accessible turns, easier than pulling a bundle but
harder than unbolting a plate stack. The choice of a spiral is therefore a choice
about the fluid: a clean high-pressure duty goes to shell-and-tube, a clean
close-approach duty to plate-and-frame, and a dirty, viscous, or fibrous duty that
would plug either goes to the spiral, whose single self-scouring passage is worth its
lower area density and its fixed geometry. The vignette closes the device survey the
same way the others did, with the binding constraint deciding the class: here it is not
duty, cost, or carry-over but the fouling tendency of a difficult fluid, and the
spiral is the answer when that tendency is what governs.

\section{Pressure drop and pumping power}\pathtag{standard}

A heat exchanger that delivers heat at low pressure drop costs
little to operate. A heat exchanger that delivers the same heat at
high pressure drop requires substantial pumping power and may be
uneconomical even if its purchase cost is low. The design trade-off
between heat transfer and pressure drop is the central optimisation
of heat-exchanger design.

\engterm{The friction-pressure-drop balance}. The pressure drop in
fully-developed pipe flow is given by the Darcy-Weisbach equation
\[
\Delta p = f \frac{L}{D} \frac{\rho u^2}{2},
\]
with $f$ the Darcy friction factor (read from the Moody chart or a
correlation as a function of Reynolds number and relative roughness),
$L$ the length, $D$ the hydraulic diameter, $\rho$ the fluid density,
and $u$ the velocity. For shell-side flow, the friction factor is
correlated against the shell-side Reynolds number with corrections
for the tube layout and baffle geometry (Bell-Delaware method, Kern
method).

The shell side is harder than the tube side because the flow is not a clean pass
down a pipe but a series of crossings of the tube bundle, turned by the baffles,
with leakage paths that steal flow from the main crossflow. The tube side has one
velocity and one length; the shell side has a crossflow velocity through the bundle,
a window velocity where the flow turns around each baffle, and leakage streams
through the clearances between the baffle and the shell and between the baffle and
the tubes. The Kern method lumps all of this into a single correlation on an
equivalent shell-side hydraulic diameter and gives a quick answer good to a factor
that a margin covers; the Bell-Delaware method accounts for each leakage and bypass
stream separately with a set of correction factors and is the method design software
uses when the pressure drop must be predicted rather than bounded. The two features
that make the shell side pay attention are that its pressure drop rises steeply with
the baffle count, because each baffle forces another crossing and another turn, and
that leakage flow does no useful heat transfer, so a shell with sloppy clearances
loses both duty and pressure to streams that bypass the bundle. The design lever is
the baffle spacing: closer baffles raise the crossflow velocity and the film but
raise the pressure drop faster, so the spacing is set where the shell-side drop
lands in its target band, typically between a fifth and a whole shell diameter, the
shell-side counterpart of the tube-side velocity choice.

The pumping power required to push a fluid through the exchanger is
$\dot{W} = \dot{V} \Delta p / \eta_p$, with $\dot{V}$ the volumetric
flow rate and $\eta_p$ the pump efficiency. The pumping power scales
as $u^3$ for fixed pipe geometry (since $\Delta p \propto u^2$ and
$\dot{V} \propto u$). Doubling the flow velocity quadruples the
pressure drop and increases the pumping power by a factor of eight.

\engterm{Heat-transfer versus pressure-drop trade-off}. The heat-
transfer coefficient inside a pipe scales as $h \propto u^{0.8}$ for
turbulent flow (Dittus-Boelter); the pressure drop scales as $u^{2}$.
The ratio $h/\Delta p$ scales as $u^{-1.2}$, so lower velocities
give better heat-transfer-per-pressure-drop. The optimisation
constraint is the required heat-transfer area, which scales as
$1/h$; lower velocities require more area, and the area cost grows
faster than the pumping-power saving below some critical velocity.
The optimal velocity for engineering services is typically in the
range $1$ to $3\,\text{m/s}$ for liquids in tubes and $10$ to $30\,
\text{m/s}$ for gases.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-pressure-drop-tradeoff.tex}

The minimum is set by the balance of the two exponents. Writing the
total annualised cost as $C(u) = k_{\text{cap}} u^{-0.8} +
k_{\text{pump}} u^{3}$ and setting $dC/du = 0$ gives the optimum
velocity $u^\star = [0.8 k_{\text{cap}}/(3 k_{\text{pump}})]^{1/3.8}$.
The exponent $1/3.8$ flattens the dependence on the cost ratio: a
tenfold error in the assumed cost ratio moves the optimum velocity by
less than a factor of two. The routine
\texttt{pressure\_drop\_optimum.py} evaluates both the scan and the
analytic optimum and confirms that they agree.

The optimisation may be done explicitly through a cost function that
combines capital cost (a function of area) and operating cost (a
function of pumping power, integrated over the service lifetime).
Process-design software embeds the optimisation as a built-in
function; hand calculations use rules of thumb (e.g., the Kern method's
recommendation that the shell-side pressure drop be a fraction of the
tube-side, with both in a target range).

\engterm{The Reynolds analogy and why heat and friction rise together}. The
reason the heat-transfer coefficient and the pressure drop climb together with
velocity is not a coincidence of two correlations but a single piece of physics:
both are carried by the same turbulent eddies. Momentum is transported to the wall
by the eddies and shows up as wall shear, which is the friction; heat is
transported to the wall by the same eddies and shows up as the film coefficient.
The Reynolds analogy states the connection as $\mathrm{St} \approx f/8$, relating
the Stanton number $\mathrm{St} = h/(\rho u c_p)$, a dimensionless heat-transfer
coefficient, to the friction factor $f$. The Chilton-Colburn form corrects it for
fluids whose momentum and heat diffuse at different rates, $\mathrm{St}\,
\mathrm{Pr}^{2/3} = f/8$, with the Prandtl number $\mathrm{Pr}$ carrying that
ratio. The analogy has a hard consequence for design: any device that raises the
film coefficient by roughening the surface, twisting the flow, or inserting
turbulators raises the friction in near-proportion, so there is no free lunch in
which heat transfer is enhanced without paying pressure drop. The exception that
proves the rule is a device that increases area rather than turbulence, a fin or a
larger tube count, which raises the conductance without raising the local friction,
which is why extended surface and not turbulence promoters is the first move on a
low-coefficient side. An enhancement is worth its pressure-drop cost only when it
raises the film on the governing side by more than it raises the pumping there, a
test the Reynolds analogy makes quantitative.

\engterm{Allocating the pressure-drop budget between the two sides}. A design is
given a pressure-drop allowance on each stream, and spending each allowance fully
is what extracts the most film for the pumping the plant has agreed to pay. An
allowance left unspent is a film coefficient left on the table, because the velocity
could have been raised until the drop reached its limit, so a good design runs each
side close to its allowance rather than well under it. The two sides are set by
different levers: the tube side by the number of passes and the tube count, which
fix the velocity for a given flow, and the shell side by the baffle spacing and cut,
which fix the crossflow velocity. When the two allowances cannot both be met, the
governing side, the one whose film limits the coefficient, gets the velocity and the
other side is relaxed, because raising the velocity on the non-governing side buys
almost no duty while spending pressure drop. The Kern rule that the shell-side drop
be a fraction of the tube-side is a starting split, not a law; the real allocation
follows the resistance stack, spending pumping where the film governs and conceding
it where the film is already comfortable, which is the same logic the kerosene
cooler's term-by-term network applied to the coefficient now applied to the
pressure-drop budget.

\begin{mastery}
The pressure-drop-versus-heat-transfer trade has a compact figure of merit that
lets two surfaces be compared without a full design, the area goodness factor and
its cousin the volume goodness factor. Define the heat-transfer power per unit area
as the Colburn factor $j = \mathrm{St}\,\mathrm{Pr}^{2/3}$ and the friction power
per unit area through $f$; the ratio $j/f$ measures how much heat a surface moves
per unit of pumping it costs, and a surface with a higher $j/f$ at the design
Reynolds number is the better choice on pumping grounds. The subtlety is that $j/f$
alone favours smooth, low-turbulence surfaces, which move heat efficiently per
watt of pumping but need large area, so a second criterion, the compactness or heat
transfer per unit volume, pulls the other way toward the small-channel, high-$j$
surfaces that pack area densely at the cost of a worse $j/f$. The design lives at the
balance of the two: an aircraft or a cryogenic plant that pays dearly for volume
and weight accepts a poor $j/f$ to get a compact core, while a stationary process
plant with cheap floor space and expensive electricity picks the high-$j/f$ surface
and lets the exchanger be large. The two goodness factors are the quantitative form
of the recurring choice between a compact high-pumping unit and a generous
low-pumping one, and they let a designer rank candidate surfaces from published
$j$ and $f$ curves before committing to a geometry, which is how compact-exchanger
selection is actually done.
\end{mastery}

\section{Thermal energy storage}\pathtag{standard}

Many thermal systems require energy storage: solar thermal plants
store daytime collected energy for night-time use, district heating
networks store off-peak heat for peak use, industrial processes
store heat between batch operations. Three working storage modes:

\engterm{Sensible heat storage}. Heat is stored as a temperature rise
in a material (water, oil, salt, concrete, rock). The stored energy
is $Q = m c \Delta T$. Water at moderate temperatures ($30$ to $90\,
\degC$) is the cheapest medium per unit stored energy; molten salt
(typical mixtures of sodium and potassium nitrate) is the working
high-temperature ($300$ to $560\,\degC$) medium for concentrated
solar power plants.

\engterm{Latent heat storage}. Heat is stored in a phase change at a
fixed temperature: ice-water for cooling, paraffin waxes for moderate
temperatures, salt hydrates for moderate-to-high temperatures. Latent
storage offers high energy density per unit volume but at a fixed
temperature, complicating systems where the load temperature varies.
Phase-change materials (PCMs) for building thermal mass and for
thermal storage in textiles are a substantial commercial product
category.

\engterm{Thermochemical storage}. Heat drives an endothermic chemical
reaction (e.g., dehydration of a salt hydrate); the reaction stores
the heat as chemical free energy and releases it on the reverse
reaction. Thermochemical storage offers very high energy density and
indefinite storage duration at ambient temperature, but the working
systems remain mostly in research as of 2026. The energy-density
advantage is worth a number. A salt hydrate such as magnesium chloride
hexahydrate releases on the order of $2.5\,\text{GJ}$ per cubic metre of
material when it is dehydrated and rehydrated across its full reaction,
against the $150\,\text{MJ/m}^3$ a $30\,\si{\kelvin}$ water tank holds and
the $180\,\text{MJ/m}^3$ of a paraffin latent store, so the thermochemical
route packs roughly fifteen times the water tank's energy into the same
volume. The indefinite duration follows from the storage mechanism: once
the reactant is dried and the two products are kept apart, no energy leaks
away, because the store holds its charge as separated chemical species
rather than as a temperature above ambient that a wall conductance bleeds
off. A thermochemical store therefore has no standing-loss time constant of
the kind the water tank showed, which is what makes it the candidate for
true seasonal storage where a sensible store's months-long time constant is
still too short. The reasons it stays in research are the engineering of the
reactor, not the thermodynamics: the reaction rate is limited by heat and
mass transfer into a bed of solid reactant that sinters, swells, and loses
permeability over cycles, the corrosive salts attack containment, and the
charge and discharge power, set by that bed transport, falls far short of
the energy the bed holds, which is the same energy-versus-power split the
sensible and latent stores showed, here at its most severe. A store that
holds a season of heat and can deliver it only slowly needs a reactor
designed for power the way a latent store needs its buried heat exchanger.

\engterm{Working numbers}. A $1\,\text{m}^3$ tank of hot water at $80\,
\degC$ (with $20\,\degC$ surroundings) stores $\sim 70\,\text{kWh}$
of thermal energy as sensible heat. A $1\,\text{m}^3$ tank of phase-
change material melting at $30\,\degC$ with latent heat $200\,\text{kJ/kg}$
stores $\sim 50\,\text{kWh}$ as latent heat. A $1\,\text{m}^3$ block
of concrete at $60\,\degC$ in $20\,\degC$ ambient stores $\sim 24\,
\text{kWh}$. The choice between modes is set by the temperature range,
the duty cycle, and the cost. The dataset
\texttt{storage\_media.csv} collects the density, specific heat,
latent heat, and usable temperature swing behind these numbers.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-storage-comparison.tex}

\subsection{The power a store can deliver, not just the energy it holds}

A storage medium is chosen for two numbers that pull in different directions: the
energy it holds and the power it can charge or discharge, and confusing the two
undersizes real systems. The energy is the medium's heat capacity times its usable
temperature swing, a property of the bulk; the power is the rate at which heat can
be moved into or out of that bulk, set by a heat exchanger between the medium and
the working fluid. A large sensible store of low conductivity, a rock bed or a
concrete block, holds abundant energy but delivers it slowly because heat must
conduct from the interior of each rock to its surface, so its discharge power is
limited by the internal conduction time constant $\tau \sim L^2/\alpha$, with $L$
the characteristic half-size of a rock and $\alpha$ its thermal diffusivity. Take a
$0.1\,\text{m}$ rock of $\alpha = 1\times 10^{-6}\,\text{m}^2/\text{s}$: the
conduction time is $\tau \sim (0.05)^2/10^{-6} = 2500\,\text{s}$, about forty
minutes, so a rock bed cannot be discharged in seconds no matter how much air is
blown through it. A water store sidesteps this because water is pumped through the
exchanger and mixes, so its power is set by the exchanger area and the pump, not by
conduction through a solid. The design rule is to size the medium for the energy and
the store's heat exchanger for the power, and to check that the medium can release
its heat as fast as the exchanger can carry it away, which small-particle packed
beds and finned phase-change modules exist to guarantee.

\subsection{Self-discharge and the standing-loss time constant}

A store loses heat to its surroundings whether or not it is being used, and the
standing loss sets how long the store holds a useful charge. Model a tank of
stored energy $E = m c\,\Delta T$ losing heat through an overall wall conductance
$U_{\text{wall}} A_{\text{wall}}$ to an ambient at $T_\infty$; the temperature
above ambient decays exponentially with a time constant $\tau_{\text{loss}} =
m c/(U_{\text{wall}} A_{\text{wall}})$. Take a $50\,\text{m}^3$ water tank
($m = 5\times 10^4\,\text{kg}$, $c = 4180\,\text{J}/(\text{kg}\cdot\si{\kelvin})$)
with $0.2\,\text{m}$ of insulation ($k = 0.04\,\si{\watt\per\meter\per\kelvin}$)
over a wall area of about $75\,\text{m}^2$, giving a wall conductance per unit area
of $k/L = 0.04/0.2 = 0.2\,\si{\watt\per\meter\squared\per\kelvin}$ and a total
$U_{\text{wall}} A_{\text{wall}} = 0.2\times 75 = 15\,\text{W}/\si{\kelvin}$. The
time constant is $\tau_{\text{loss}} = 5\times 10^4\times 4180/15 = 1.4\times 10^7\,
\text{s}$, about $160\,\text{days}$, so the tank loses a fixed fraction of its
above-ambient charge every $160\,\text{days}$ and holds heat well for a season.
Scale the same insulation down to a $1\,\text{m}^3$ domestic tank and the surface-
to-volume ratio rises, the time constant falls to a few days, and the store is a
day-to-day buffer rather than a seasonal one. The lesson the time constant carries
is that large stores self-discharge slowly because their volume grows faster than
their surface, which is the physics behind seasonal pit stores tens of metres
across: at that scale the surface-to-volume ratio is small enough that summer heat
survives to winter with a loss the design can absorb.

\subsection{Latent storage and the two-temperature trap}

Latent storage packs more energy into less volume by storing heat in a phase change
at a fixed temperature, and the fixed temperature is both its strength and its
trap. A phase-change material melting at $T_m$ absorbs its latent heat at that one
temperature, so a $1\,\text{m}^3$ store of a wax with latent heat $200\,\text{kJ/kg}$
and density $900\,\text{kg/m}^3$ holds $900\times 200 = 1.8\times 10^5\,\text{kJ} =
50\,\text{kWh}$ across a negligible temperature swing, against the $34\,\text{kWh}$
a water tank holds across a $30\,\si{\kelvin}$ swing. The trap is that heat flows
into or out of the phase-change store only when the working fluid is on the correct
side of $T_m$, so a store melting at $30\,\degC$ charges only from a fluid warmer
than $30\,\degC$ and discharges only to a fluid cooler than $30\,\degC$, and the
approach temperature that drives the transfer eats into the swing the working fluid
has to spend. A second difficulty is that most phase-change materials conduct heat
poorly in the solid phase, so as the store discharges, a growing solid layer
insulates the still-molten core, and the discharge power falls through the cycle
unless the store is built with a conductive matrix, fins, or encapsulation in small
modules to keep every parcel of material close to a heat-transfer surface. The
design of a latent store is therefore mostly the design of the heat exchanger
buried in it, which is why the storage-media table lists a usable swing and a
conductivity alongside the latent heat: the latent heat sets the energy, the
conductivity and the exchanger set the power, and a store chosen on latent heat
alone will hold its charge and refuse to give it up at the rate the load demands.

\begin{mastery}
A sensible-heat water store is worth more than its bulk energy content
suggests when it holds thermal stratification. Hot water is less dense
than cold, so a well-designed tank keeps a sharp thermocline between a
hot layer at the top and a cold layer at the bottom, and the store
delivers water at the full hot temperature until the thermocline
reaches the outlet. A fully mixed tank of the same size delivers water
at the average temperature throughout, which falls steadily and may
drop below the load's required supply temperature long before the
energy is spent. The figure of merit is the degree of stratification,
degraded by inlet jet mixing, by conduction across the thermocline, and
by standing heat loss through the walls. Diffuser inlets that spread
the incoming flow at low velocity, tall narrow tank geometries that
limit the thermocline area, and short cycle times all preserve
stratification. The same physics governs the pit thermal stores of
district networks, where the thermocline can be metres thick and still
small against a tank tens of metres deep.
\end{mastery}

\section{District heating and cooling}\pathtag{mastery}

A district energy system centralises the production of heating and
cooling for a neighbourhood, a campus, or a city and distributes the
thermal energy through insulated pipes. The pattern is most
established in Scandinavia (Copenhagen's district-heating network
covers $98\,\%$ of buildings), Eastern Europe, parts of East Asia,
and increasingly in North American and Western European
metropolitan areas.

\engterm{District-heating networks} typically operate at one of three
temperature classes:

\begin{itemize}[leftmargin=*]
  \item First-generation: high-temperature steam ($\sim 180\,\degC$).
    Mostly retired.
  \item Second-generation: high-temperature water ($120$ to $150\,\degC$).
    Found in older systems.
  \item Third-generation: medium-temperature water ($80$ to $120\,\degC$).
    The dominant current technology.
  \item Fourth-generation: low-temperature water ($50$ to $70\,\degC$).
    The current build-out technology, suited to heat-pump-driven
    sources and low-temperature waste heat.
  \item Fifth-generation: very-low-temperature networks ($10$ to $25\,
    \degC$, near ambient). Each building has its own heat pump that
    raises or lowers the network temperature locally. Suited to
    deep geothermal, mine water, and ambient-loop topologies.
\end{itemize}

The trend is toward lower temperatures because lower-temperature
networks have lower distribution heat losses, can be supplied by
waste heat from data centres and industrial processes, and are
compatible with electric heat-pump sources.

\engterm{District-cooling networks} circulate chilled water (typical
supply $5$ to $7\,\degC$, return $13$ to $15\,\degC$) from a central
chilling plant to consumer buildings. Economies of scale in the
chillers (large absorption or vapour-compression units) and the
ability to use seawater or river-water cooling make district cooling
substantially more efficient than building-by-building air
conditioning in dense urban areas. The temperature difference the
network is designed for, the delta-T between supply and return, is the
number that fixes the pipe size and the pumping cost, because the flow a
network must circulate to carry a given cooling load is inversely
proportional to it. A network built for an $8\,\si{\kelvin}$ rise
($7\,\degC$ supply, $15\,\degC$ return) carries half the flow of one
built for a $4\,\si{\kelvin}$ rise at the same duty, and since pumping
power scales with the cube of the flow, the wide-delta-T network pumps
at roughly an eighth of the power. The chronic fault of real
district-cooling systems is the low-delta-T syndrome: consumer
substations that return water warmer than design (a fouled coil, an
oversized or hunting control valve, a bypass left open) shrink the
network delta-T, force the flow up to hold the duty, and drive the
pumping cost and the pipe velocity past their design points, the
cooling counterpart of the high-return-temperature fault the substation
diagnosis exercise reads on the heating side. The central plant's
efficiency compounds the network gain: a large water-cooled chiller
reaching a coefficient of performance near five against a cooling tower
or seawater sink beats the air-cooled split units of individual
buildings, which reject to hot ambient air at a coefficient nearer
three, so the district plant delivers the same cooling for roughly half
the electricity before the distribution loss is counted. The
distribution loss runs the other way from district heating: a chilled
network gains heat from warmer surroundings, so its insulation fights
an inward leak, and the low supply temperature makes that leak a larger
fraction of the small delta-T the network carries, which is why chilled
mains are insulated and often buried below the summer-warm surface soil.

\engterm{Network design}. The hydraulic design of a district network
follows the steady-flow pressure-drop calculation of section 6.4
applied to a graph: each segment has a friction-pressure-drop, each
consumer takes a flow, and the central plant must deliver the total
pressure at the worst-case demand. Modern networks are designed for
variable flow with consumer-side control valves; the central pumps
are variable-speed-driven to deliver the required pressure at each
operating point. Volume IX develops the control side.

\engterm{Thermal storage in district networks}. Central thermal-
storage tanks (typically large pits or steel tanks of $10^4$ to $10^5
\,\text{m}^3$) allow the central plant to operate at off-peak times
or to ride through interruptions in waste-heat supply. Marstal,
Denmark's $60{,}000\,\text{m}^3$ pit thermal energy storage is a
canonical example, providing roughly $400$ days of summer collected
solar heat to a winter heat-pump-driven district heating system.

\begin{mastery}
A plant with many hot streams to cool and many cold streams to heat
must decide how much of the heating can be supplied by the cooling, so
that external utilities carry only the remainder. Pinch analysis fixes
that quantity before any exchanger is placed.
Sort every stream into temperature intervals and, within each interval,
add the heat available from all hot streams to form the hot composite
curve and the heat required by all cold streams to form the cold
composite curve. Each curve is temperature against cumulative enthalpy.
Sliding the cold curve horizontally until its closest vertical approach
to the hot curve equals a chosen minimum approach $\Delta T_{\min}$ fixes
three numbers at once: the horizontal overlap is the heat internal
exchangers can recover, the cold-end overshoot of the hot curve is the
minimum cold utility, and the hot-end overshoot of the cold curve is the
minimum hot utility. Figure \ref{fig:vol04ch06:composite-curves} shows
the construction. The point of closest approach is the pinch, and it
splits the process into a region above, which is a net heat sink, and a
region below, which is a net heat source. Three design rules follow and
together guarantee the minimum-utility targets: do not transfer heat
across the pinch, do not use cold utility above the pinch, and do not use
hot utility below the pinch. Choosing $\Delta T_{\min}$ trades capital
against energy in the same shape the single-exchanger area-versus-driving-
difference trade-off takes: a small approach recovers more heat but
needs more area, and the optimum sits where the marginal area cost meets
the marginal utility saving. The targets are known before a single unit
is sized, which lets the engineer judge a flowsheet against the
thermodynamic best case rather than against an arbitrary baseline.
\end{mastery}

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-composite-curves.tex}

\section{Failure: fouling, corrosion, tube leaks}\pathtag{core}

The most common heat-exchanger failure modes are not catastrophic
ruptures but progressive degradations that reduce performance and
require maintenance. The chapter's failure section covers fouling,
corrosion, and tube leaks.

\subsection{Fouling}

\engterm{Fouling} is the accumulation of deposits on the heat-
transfer surfaces, reducing the effective heat-transfer coefficient.
Six fouling mechanisms occur in engineering services:

\begin{itemize}[leftmargin=*]
  \item Particulate fouling: suspended particles (silt, rust, scale)
    in the fluid deposit on the heat-transfer surface.
  \item Crystallisation fouling: dissolved species crystallise on the
    surface as their solubility falls with temperature (cooling
    water depositing calcium carbonate, brackish water depositing
    calcium sulphate, ammonia/water systems depositing salts).
  \item Biological fouling: bacteria, algae, and biofilms grow on
    the surface. Found in seawater cooling, in cooling-tower water,
    and in any system with stagnant or low-temperature water and
    nutrients.
  \item Chemical-reaction fouling: the fluid undergoes a slow
    chemical reaction that produces a solid deposit. Polymerisation
    of hydrocarbons on hot surfaces, coking of crude-oil
    distillation columns, scale formation in steam boilers from
    water hardness.
  \item Corrosion-product fouling: the heat-exchanger material
    corrodes, producing oxide deposits that build up on the surface.
  \item Freezing fouling: a process fluid near its freezing point
    deposits a solid layer on the cold surface (wax in crude oil,
    ice in food-processing freezers).
\end{itemize}

The engineering accounting for fouling is the \engterm{fouling
resistance} $R_f$, added in series to the clean overall heat-transfer
coefficient:
\[
\frac{1}{U_{\text{dirty}}} = \frac{1}{U_{\text{clean}}} + R_f.
\]
Standard fouling resistances are tabulated in TEMA standards and
the Tubular Exchanger Manufacturers Association handbook. For
cooling water, $R_f \approx 0.0002$ to $0.001\,\text{m}^2\cdot\si{
\kelvin}/\text{W}$; for crude oil, $0.0009$ to $0.002$; for distillation
products, $0.0002$ to $0.0007$. The fouling resistance is added on
both sides of the exchanger when both are subject to fouling.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-thermal-network.tex}

A typical clean overall coefficient of $500\,\text{W}/(\text{m}^2
\cdot\si{\kelvin})$ has $1/U_{\text{clean}} = 0.002\,\text{m}^2\cdot
\si{\kelvin}/\text{W}$. A fouling resistance of $0.001$ on each side
adds $0.002$, doubling the total resistance and halving the heat-
transfer rate. Designers typically over-size the exchanger by a
factor that allows for fouling at the end of a cleaning cycle (six
months to two years between cleanings, depending on service). The
tabulated fouling resistances used in design appear in
\texttt{fouling\_resistances.csv}; the routine
\texttt{overall\_coefficient.py} assembles the series network and
returns the clean and dirty coefficients.

\input{volumes/04-energy/ch06-heat-exchangers-thermal-systems/figures/fig-fouling-decline.tex}

The fouling layer does not grow without bound. Deposition lays material
down while the shear of the flowing fluid scours it away, and the two
rates come into balance at an asymptotic deposit thickness. The
overall coefficient therefore falls fast at first and then levels off,
which is the curve plotted in figure
\ref{fig:vol04ch06:fouling-decline}. The dataset
\texttt{condenser\_U\_decline.csv} carries a two-cycle record of
coefficient and pressure drop against service month; the residual loss
after the second cleaning is the hard scale that mechanical cleaning
does not remove, and tracking it is how a plant decides when retubing
beats further cleaning.

\subsection{Corrosion}

Heat-exchanger corrosion modes include uniform corrosion (a slow
overall material loss), pitting (localised attack producing deep
pits, often in chloride-containing waters), galvanic corrosion
(between dissimilar metals in contact in an electrolyte), stress-
corrosion cracking (cracking under combined stress and corrosion in
specific environments), and erosion-corrosion (combined fluid impact
and corrosion, particularly at flow disturbances).

The modes differ in how they consume metal and in how they announce themselves,
which is what lets a diagnostic tell them apart. Uniform corrosion removes wall at a
roughly steady rate over the whole wetted surface, so it is the benign case a
corrosion allowance is added for: a design that specifies a $3\,\si{\milli\meter}$
allowance on a $2\,\si{\milli\meter}$-thick tube against a known uniform rate simply
runs until the allowance is spent. Pitting is the dangerous opposite, because it
concentrates the attack at a few points that penetrate the wall while the bulk
surface stays sound, so a pitted tube can perforate at a spot with almost no measured
average metal loss, which is why chloride waters are watched by inspection for
localised indications rather than by average thickness. Galvanic corrosion needs
two metals of different electrode potential in electrical contact in an electrolyte:
the less noble metal becomes the anode and corrodes preferentially, so a
carbon-steel tubesheet paired with a copper-alloy tube in seawater sacrifices the
steel, and the accelerating factor is the area ratio, because a small anode feeding
a large cathode corrodes fast. Stress-corrosion cracking needs three things at once,
a susceptible alloy, a specific environment, and a sustained tensile stress, and it
produces fine cracks with little general metal loss, which is why it is caught by
crack-imaging inspection at high-stress locations rather than by thickness gauging.
Erosion-corrosion strips the protective oxide film mechanically where the flow turns
or accelerates, exposing fresh metal that corrodes and is stripped again, so it
localises at inlet nozzles, tube ends, and U-bends and is fought by lowering the
local velocity, rounding the flow path, or upgrading the surface. The diagnostic
that distinguishes them reads the pattern of loss against the environment: general
thinning points to uniform attack, deep isolated pits to pitting, preferential loss
of one metal at a bimetallic junction to galvanic action, fine branching cracks at
stress risers to cracking, and grooved loss at flow disturbances to
erosion-corrosion. Each mode has a matched defence, so naming the mode is the step
that turns a leak into a repair.

Material selection is the primary defence:

\begin{itemize}[leftmargin=*]
  \item Carbon steel: cheap, suitable for steam, condensate, treated
    cooling water in non-corrosive service.
  \item Stainless steel (304, 316): general corrosion-resistant,
    vulnerable to chloride stress-corrosion cracking above $\sim 60\,
    \degC$ in chloride environments.
  \item Copper alloys: good for water service, vulnerable to
    ammonia corrosion.
  \item Titanium: outstanding corrosion resistance in seawater and
    chloride environments, expensive but cost-effective for severe
    service.
  \item Inconel, Monel, Hastelloy: specialty alloys for high-
    temperature corrosion and severe chemical environments.
\end{itemize}

The selection is an ordered walk up the cost ladder, stopping at the first material
whose corrosion rate the service can tolerate over the design life. Carbon steel is
tried first because it is cheapest, and it is accepted wherever a corrosion allowance
of a millimetre or two of extra wall, spent at the known uniform rate, outlasts the
design life; a treated-cooling-water or clean-steam service is the usual home. The
step to stainless is taken when uniform attack on steel is too fast, but stainless
carries its own trap, chloride stress-corrosion cracking, which appears above roughly
$60\,\degC$ in chloride-bearing water and is not cured by adding wall, because
cracking penetrates with almost no general metal loss, so a hot chloride service jumps
past ordinary stainless to a duplex grade or to titanium rather than up-gauging the
304 or 316 tube. Copper alloys hold the water-service niche between steel and the
exotics, prized for their resistance to biofouling and general seawater corrosion, but
ruled out wherever ammonia is present, which cracks them, or where the velocity
exceeds the alloy's erosion limit, near $2\,\text{m/s}$ for the common
copper-nickels, above which the flowing water strips the protective film faster than
it reforms. Titanium is the answer for severe seawater and chloride duty because it
carries a self-healing oxide film immune to pitting and erosion across the full
velocity range, and although its purchase price is several times the copper alloys,
the life-cycle comparison of the seawater worked example is what justifies it: a tube
that never returns to the shop beats a cheaper tube replaced twice over the plant
horizon once the outage cost is counted. The specialty alloys sit at the top of the
ladder for high-temperature oxidation and aggressive chemical environments where even
titanium fails, and they are specified only when the service leaves no cheaper choice,
because their cost per kilogram is an order above stainless. The discipline is to name
the corrosion mode the service will present, uniform, pitting, cracking, galvanic, or
erosion, and to select the first material up the ladder that defeats that specific
mode, rather than reaching for an expensive alloy as a blanket insurance the service
may not need.

\subsection{Tube leaks}

A tube leak in a shell-and-tube exchanger allows mixing of the two
fluids. The consequences depend on the service: in a steam-
condensate-to-cooling-water exchanger, a leak introduces cooling
water into the condensate, contaminating it; in a chemical-process
exchanger, the leak may introduce a reactive species, causing a
runaway, a corrosion attack, or contamination. In nuclear power
plant steam generators, a tube leak releases primary-coolant
radioactivity into the secondary side and into the environment.

The direction of a leak is set by the pressure difference, which is what makes some
leaks a nuisance and others a hazard. Fluid flows from the higher-pressure side to
the lower, so a designer arranges the pressures to keep the leak flowing the safe
way whenever the service allows it: in a cooling-water condenser the cooling water
is held above the condensate pressure so a small tube leak sends clean cooling water
into the condensate rather than drawing air and contamination inward, while in a
service where cross-contamination must be prevented in both directions the answer is
a double-tubesheet or double-wall construction with a monitored gap between the two
walls, so a leak in either wall shows up in the gap before the two fluids can meet.
The consequences scale with what the two fluids do to each other on contact: two
compatible fluids give a contamination that is caught by analysis and cleaned up,
whereas two reactive fluids give a runaway, a fire, or a pressure excursion the
moment they mix, which is why exchangers between reactive streams carry the
double-wall construction and a leak-detection gap as a matter of course.

The diagnostic for tube leaks proceeds from cheapest and most sensitive to most
invasive. Chemical analysis of the lower-pressure side for the contaminant species
detects a leak in service without shutting down, because a trace of the
higher-pressure fluid appearing where it should not is unambiguous evidence; in a
nuclear steam generator the primary-side activity monitors on the secondary side are
this same principle run continuously with a radioactive tracer already present.
Once a leak is confirmed, the failed tube is located by isolating the exchanger and
pressure-decay testing sections, or by helium leak detection, which fills one side
with helium and sniffs the other with a mass-spectrometer detector sensitive to
parts per million. Eddy-current inspection then images the tube walls to map the
wall loss and crack indications that reveal not just which tube leaked but by what
mechanism, so the same inspection that finds the leak also names the corrosion mode
that caused it and guides whether to plug a few tubes or retube the bundle. The
sequence, monitor in service, locate on shutdown, image to diagnose, is the
discipline that keeps a small leak from becoming an unplanned outage.

\subsection{The diagnostic principle for heat-exchanger failures}

The maintenance discipline for heat exchangers tracks several
indicators:

\begin{itemize}[leftmargin=*]
  \item Overall heat-transfer coefficient: declining $U$ over time
    indicates fouling.
  \item Pressure drop: rising $\Delta p$ over time indicates fouling
    or partial blockage.
  \item Outlet temperature anomalies: a stream's outlet temperature
    drifting from design indicates a flow or capacity change.
  \item Vibration: tube-bundle vibration above the design level
    indicates flow-induced excitation, often a precursor to tube
    failure.
  \item Chemical-composition drift: process-side chemistry changing
    in ways consistent with leakage from the other side.
\end{itemize}

The asset-management decision is between cleaning (restoring the
clean condition), retubing (replacing failed tubes), and
replacement. Each has a cost; the working maintenance manager
balances the cost of cleaning against the lost-throughput cost of
operating with degraded performance.

\begin{mastery}
The vibration indicator in the diagnostic list is not a nuisance signal but the
early warning of a failure mode that can destroy a tube bundle in hours, and it has
its own design theory. Shell-side crossflow past a tube sheds vortices at a
frequency $f_s = \mathrm{St}\,u/d_o$, with $\mathrm{St}$ the Strouhal number (near
$0.2$ for a tube in crossflow), $u$ the crossflow velocity, and $d_o$ the tube outer
diameter. Each tube also has natural frequencies of its own, set by its length
between baffle supports, its material, and its mass per unit length including the
fluid it carries and displaces. When the shedding frequency approaches a tube
natural frequency the tube resonates, and the amplitude grows until the tube either
fatigues at the baffle holes or clashes with its neighbours and wears through, which
is the mid-span or baffle-edge failure a bundle autopsy shows. A second and more
dangerous mechanism is fluid-elastic instability: above a threshold crossflow
velocity the coupling between a tube's motion and the flow field feeds energy into
the vibration faster than damping can remove it, and the whole array goes unstable
at once rather than one tube resonating, a runaway that is designed against by
keeping the crossflow velocity below the critical value set by a mass-damping
parameter. The design levers are the baffle spacing, which sets both the unsupported
span (and thus the natural frequency) and the crossflow velocity, and the entrance
region, where the inlet nozzle drives the highest local velocity and where most
vibration failures start. A bundle that passes its thermal and pressure-drop checks
can still fail vibration, so the vibration analysis is a separate gate, and a rising
measured vibration in service is the field signature that the bundle is running too
close to a shedding resonance or the instability threshold.
\end{mastery}

\begin{archetype}[Optimisation under transport-resistance trade-off]
A heat exchanger's design is the optimisation of heat transfer
against pressure drop against capital cost against fouling. Each
geometric or operational choice (tube diameter, tube length, baffle
spacing, flow velocity, materials) shifts the four costs differently.
The discipline is to identify the dominant constraint (fouling,
capital, pumping power, or pressure-drop limit) and design against
it; iterating once the dominant constraint is known is far more
productive than attempting a global multi-objective optimisation.
The same archetype recurs in mass-transfer column design (chapter~7),
in process flowsheet design (Vol~V), and in any transport-limited
engineering system.
\end{archetype}

\begin{estimation}
\textbf{Estimate the heat-transfer area required for a $1\,\text{MW}$
shell-and-tube condenser using cooling-tower water.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
Steam at $50\,\degC$ saturation temperature condenses on the tube
side. Cooling water enters at $25\,\degC$ and leaves at $35\,\degC$.
The temperature differences at the two ends are $\Delta T_1 = 50 -
35 = 15\,\si{\kelvin}$ and $\Delta T_2 = 50 - 25 = 25\,\si{\kelvin}$.
The LMTD is $(\Delta T_1 - \Delta T_2)/\ln(\Delta T_1/\Delta T_2) =
(15 - 25)/\ln(15/25) \approx 19.6\,\si{\kelvin}$.

For a steam condenser with cooling water, typical overall heat-
transfer coefficients are $U = 1500$ to $3000\,\text{W}/(\text{m}^2
\cdot\si{\kelvin})$. Take $U = 2000\,\text{W}/(\text{m}^2\cdot\si{
\kelvin})$.

The required area is $A = \dot{Q}/(U \Delta T_{\text{LMTD}}) = 10^6/(
2000 \times 19.6) \approx 25.5\,\text{m}^2$.

For tubes of $25\,\si{\milli\meter}$ outer diameter, the surface area
per metre is $\pi D = 0.079\,\text{m}^2/\text{m}$, so the total tube
length required is $25.5/0.079 = 320\,\text{m}$. A $2\,\text{m}$-long
shell with $160$ tubes provides this. The shell diameter, with a
typical tube pitch of $32\,\si{\milli\meter}$ in a triangular array,
is roughly $0.5\,\text{m}$.

The estimate gives an exchanger of roughly $0.5\,\text{m}$ shell
diameter and $2\,\text{m}$ length, with $160$ tubes. Published
designs for $1\,\text{MW}$ steam condensers cluster in this range
\cite{text:cengel-boles}. The postmortem: the dominant
sensitivity is the assumed $U$ (factor of two range across typical
designs); the LMTD calculation is well-bounded. Cleaning interval
and fouling allowance would adjust the design point by perhaps
$30\,\%$, within the same order of magnitude.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Design a heat exchanger for a specified duty and
verify with software]
The reader is given a heat-transfer duty (e.g., cooling a $10\,\text{kg/s}$
stream of hot oil from $120\,\degC$ to $60\,\degC$ using cooling water
from $25\,\degC$ to $40\,\degC$), designs a shell-and-tube exchanger
by hand using the LMTD method, then verifies with a process-
simulation package.

\textbf{Track}. Design plus software verification.
\textbf{Hazard class}: low (paperwork and software).

\textbf{Hand design}.

\begin{itemize}[leftmargin=*]
  \item Heat balance: compute the required heat-transfer rate from
    the hot-side temperatures and mass flow.
  \item Cold-side flow: compute the cooling-water mass flow from the
    cold-side temperature change.
  \item LMTD: compute the counter-flow LMTD and any correction factor
    for the chosen geometry.
  \item Overall U-value: estimate from typical values for the service
    and the chosen materials. State the fouling resistances assumed.
  \item Area: compute from $\dot{Q} = U A F \Delta T_{\text{LMTD}}$.
  \item Geometry: specify tube size, length, number of tubes, shell
    diameter, baffle spacing, number of passes.
  \item Pressure drop: compute on both sides and verify acceptable.
\end{itemize}

\textbf{Software verification}. Use a process-simulation package
(Aspen EDR, HTRI, ChemCAD, or an open-source equivalent) to
reproduce the design point. Compare the software's $U$, area, and
pressure drops to the hand calculation.

\textbf{Reflection ($1{,}500$ to $2{,}500$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
  \item Where did the hand calculation and the software agree?
    Where did they differ?
  \item Which assumption in the hand calculation was the most
    sensitive (most likely to drive the discrepancy)?
  \item What over-design margin would the reader specify in the
    final design, and on what basis?
\end{itemize}

\textbf{Deliverable}. The hand-design report, the software output,
the comparison, and the reflection.

\textbf{Time}. Six to ten hours of hand design, two to four hours of
software work, four to six hours of write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A counter-flow heat exchanger cools $2\,\text{kg/s}$ of hot water
from $80\,\degC$ to $40\,\degC$ using cold water entering at $20\,
\degC$ at $4\,\text{kg/s}$. Find the cold-water exit temperature, the
LMTD, and the area required for $U = 1500\,\text{W}/(\text{m}^2\cdot
\si{\kelvin})$.

\paragraph{Solution sketch.} The duty is $\dot{Q} = \dot{m}_h c_p \Delta
T_h = 2 \times 4180 \times 40 = 334\,\text{kW}$. The cold-side balance
gives $T_{c,o} = 20 + \dot{Q}/(\dot{m}_c c_p) = 20 + 334{,}400/(4
\times 4180) = 40\,\degC$. The end differences are $\Delta T_1 = 80 -
40 = 40\,\si{\kelvin}$ and $\Delta T_2 = 40 - 20 = 20\,\si{\kelvin}$,
so $\Delta T_{\text{LMTD}} = (40 - 20)/\ln(40/20) = 28.9\,\si{\kelvin}$.
The area is $A = \dot{Q}/(U\,\Delta T_{\text{LMTD}}) = 334{,}400/(1500
\times 28.9) = 7.7\,\text{m}^2$.
\end{exercise}

\begin{exercise}
Repeat the previous exercise as parallel flow. Compare the area
required and identify the second-law cost of the parallel arrangement.

\paragraph{Solution sketch.} The duty and outlet temperatures are
unchanged ($\dot{Q} = 334\,\text{kW}$, $T_{c,o} = 40\,\degC$), because
the energy balance does not depend on the flow arrangement. The end
differences become $\Delta T_1 = T_{h,i} - T_{c,i} = 80 - 20 = 60\,
\si{\kelvin}$ and $\Delta T_2 = T_{h,o} - T_{c,o} = 40 - 40 = 0\,
\si{\kelvin}$. The log-mean is indeterminate as $\Delta T_2 \to 0$ and
in fact drives the area to infinity: a parallel-flow exchanger cannot
bring the two streams to the same outlet temperature, so this exact
duty is unreachable in parallel flow. Backing the cold outlet off to,
say, $38\,\degC$ gives a finite but much larger area than the
counter-flow $7.7\,\text{m}^2$. The second-law cost is the larger
entropy generation of mixing across the steeply varying temperature
difference of the parallel arrangement; counter-flow keeps the
difference small and uniform, transferring the same heat across a
smaller average gap.
\end{exercise}

\begin{exercise}
A shell-and-tube exchanger (1 shell pass, 2 tube passes) cools
$1\,\text{kg/s}$ of oil ($c_p = 2.1\,\text{kJ}/(\text{kg}\cdot\si{
\kelvin})$) from $150\,\degC$ to $50\,\degC$ using $3\,\text{kg/s}$ of
water entering at $20\,\degC$. Find the water exit temperature, the
correction factor $F$, the LMTD, and the area for $U = 800\,\text{W}/
(\text{m}^2\cdot\si{\kelvin})$.

\paragraph{Solution sketch.} The duty is $\dot{Q} = 1 \times 2100 \times
(150 - 50) = 210\,\text{kW}$. The water outlet is $T_{c,o} = 20 +
210{,}000/(3 \times 4180) = 36.7\,\degC$. The counter-flow LMTD uses
$\Delta T_1 = 150 - 36.7 = 113.3\,\si{\kelvin}$ and $\Delta T_2 = 50 -
20 = 30\,\si{\kelvin}$: $\Delta T_{\text{LMTD,CF}} = (113.3 -
30)/\ln(113.3/30) = 62.7\,\si{\kelvin}$. The correction-factor
parameters are $P = (T_{c,o} - T_{c,i})/(T_{h,i} - T_{c,i}) = 16.7/130
= 0.13$ and $R = (T_{h,i} - T_{h,o})/(T_{c,o} - T_{c,i}) = 100/16.7 =
6.0$. From the one-shell-pass, two-tube-pass chart these values give
$F \approx 0.97$ (the low $P$ keeps $F$ near unity despite the high
$R$). The area is $A = \dot{Q}/(U F \Delta T_{\text{LMTD,CF}}) =
210{,}000/(800 \times 0.97 \times 62.7) = 4.3\,\text{m}^2$.
\end{exercise}

\begin{exercise}
An air-cooled exchanger has $A = 200\,\text{m}^2$ and $U = 50\,
\text{W}/(\text{m}^2\cdot\si{\kelvin})$. It cools $5\,\text{kg/s}$ of
oil ($c_p = 2.1\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$) entering at
$140\,\degC$ with air at $30\,\degC$ supplied at $20\,\text{kg/s}$.
Use the NTU-effectiveness method to find the outlet temperatures.

\paragraph{Solution sketch.} The capacity rates are $C_{\text{oil}} = 5
\times 2100 = 10{,}500\,\si{\watt\per\kelvin}$ and $C_{\text{air}} = 20
\times 1005 = 20{,}100\,\si{\watt\per\kelvin}$, so $C_{\min} = 10{,}500$
(oil), $C_r = 0.52$, and $\text{NTU} = UA/C_{\min} = 50 \times
200/10{,}500 = 0.95$. The cross-flow-both-unmixed correlation gives
$\varepsilon \approx 0.53$, so $\dot{Q} = \varepsilon C_{\min}(T_{h,i}
- T_{c,i}) = 0.53 \times 10{,}500 \times 110 = 610\,\text{kW}$. The oil
leaves at $140 - 610{,}000/10{,}500 = 82\,\degC$; the air leaves at $30
+ 610{,}000/20{,}100 = 60\,\degC$. The matching value is the smaller
capacity rate setting both NTU and $\dot{Q}_{\max}$. The routine
\texttt{ntu\_effectiveness.py} returns these values.
\end{exercise}

\begin{exercise}
A steam condenser uses cooling water entering at $25\,\degC$ at
$50\,\text{kg/s}$. Saturated steam at $50\,\degC$ condenses on the tube
exterior at $0.5\,\text{kg/s}$ with latent heat $2382\,\text{kJ/kg}$.
The exchanger has $U = 2500\,\text{W}/(\text{m}^2\cdot\si{\kelvin})$.
Find the water outlet temperature and the area required.

\paragraph{Solution sketch.} The duty is set by the condensing side:
$\dot{Q} = \dot{m}_{\text{steam}} h_{fg} = 0.5 \times 2382 = 1191\,
\text{kW}$. The water outlet is $T_{c,o} = 25 + 1{,}191{,}000/(50
\times 4180) = 30.7\,\degC$. The steam is at constant $50\,\degC$, so
$\Delta T_1 = 50 - 30.7 = 19.3\,\si{\kelvin}$ and $\Delta T_2 = 50 -
25 = 25\,\si{\kelvin}$: $\Delta T_{\text{LMTD}} = (19.3 -
25)/\ln(19.3/25) = 22.0\,\si{\kelvin}$. The area is $A = 1{,}191{,}000/
(2500 \times 22.0) = 21.7\,\text{m}^2$. Because one stream changes
phase, $C_r = 0$ and the LMTD method and the NTU method agree without a
correction factor.
\end{exercise}

\begin{exercise}
A plate-and-frame exchanger has area $50\,\text{m}^2$, $U = 5000\,
\text{W}/(\text{m}^2\cdot\si{\kelvin})$, and operates with
$\dot{m}_h c_{p,h} = 10\,\text{kW}/\si{\kelvin}$, $\dot{m}_c c_{p,c}
= 12\,\text{kW}/\si{\kelvin}$. Inlets are $T_{h,i} = 90\,\degC$ and
$T_{c,i} = 20\,\degC$. Compute NTU, $C_r$, the effectiveness, and the
outlet temperatures.

\paragraph{Solution sketch.} Here $C_{\min} = 10\,\text{kW}/\si{\kelvin}$
(hot side), $C_{\max} = 12\,\text{kW}/\si{\kelvin}$, $C_r = 10/12 =
0.83$, and $\text{NTU} = UA/C_{\min} = 5000 \times 50/10{,}000 = 25$.
A plate exchanger runs counter-current, so $\varepsilon = (1 - e^{-
\text{NTU}(1 - C_r)})/(1 - C_r e^{-\text{NTU}(1 - C_r)})$. With
$\text{NTU}(1 - C_r) = 25 \times 0.17 = 4.25$, the exponential is
$e^{-4.25} = 0.0143$, giving $\varepsilon = (1 - 0.0143)/(1 - 0.83
\times 0.0143) = 0.986/0.988 = 0.998$. The very high NTU drives the
effectiveness to its ceiling, which is the close approach
plate exchangers are bought for. The duty is $\dot{Q} = \varepsilon
C_{\min}(T_{h,i} - T_{c,i}) = 0.998 \times 10 \times 70 = 698\,\text{kW}$.
The hot outlet is $90 - 698/10 = 20.2\,\degC$; the cold outlet is $20 +
698/12 = 78.2\,\degC$.
\end{exercise}

\begin{exercise}
A double-pipe counter-flow heat exchanger has inner-tube diameter
$25\,\si{\milli\meter}$ and length $5\,\text{m}$. Hot water at $80\,
\degC$ flows in the inner tube at $0.2\,\text{kg/s}$; cold water at
$20\,\degC$ flows in the annulus at $0.1\,\text{kg/s}$. Estimate the
inside, outside, and overall heat-transfer coefficients, the LMTD,
and the heat-transfer rate.

\paragraph{Solution sketch.} For the inner tube the velocity is $u =
\dot{m}/(\rho A) = 0.2/(990 \times \pi/4 \times 0.025^2) = 0.41\,
\text{m/s}$, the Reynolds number is $Re = \rho u D/\mu \approx 990
\times 0.41 \times 0.025/(5 \times 10^{-4}) \approx 2.0 \times 10^4$
(turbulent), and Dittus-Boelter $Nu = 0.023 Re^{0.8} Pr^{0.4}$ with
$Pr \approx 3.5$ gives $Nu \approx 95$, so $h_i = Nu\,k/D \approx 95
\times 0.63/0.025 \approx 2400\,\si{\watt\per\meter\squared\per
\kelvin}$. The annulus flow is slower and gives a similar
order, $h_o \approx 1500\,\si{\watt\per\meter\squared\per\kelvin}$.
Neglecting wall resistance, $1/U \approx 1/2400 + 1/1500$, so $U
\approx 920\,\si{\watt\per\meter\squared\per\kelvin}$. With outlet
temperatures unknown, iterate or use NTU: the area is $A = \pi D L =
\pi \times 0.025 \times 5 = 0.39\,\text{m}^2$, $C_{\min} = 0.1 \times
4180 = 418\,\si{\watt\per\kelvin}$, $\text{NTU} = 920 \times 0.39/418 =
0.86$, $C_r = 418/836 = 0.5$, $\varepsilon \approx 0.52$, so $\dot{Q} =
0.52 \times 418 \times 60 = 13\,\text{kW}$. The exact coefficients
depend on property values and the answer is an order-of-magnitude
estimate.
\end{exercise}

\begin{exercise}
A clean shell-and-tube exchanger has $U_{\text{clean}} = 700\,\text{W}/
(\text{m}^2\cdot\si{\kelvin})$. The fouling resistances are
$0.0005\,\text{m}^2\cdot\si{\kelvin}/\text{W}$ on each side. Compute
the dirty $U$ and the fraction of the heat-transfer capacity lost
to fouling.

\paragraph{Solution sketch.} The clean resistance is $1/U_{\text{clean}} =
1/700 = 0.00143\,\text{m}^2\si{\kelvin}/\text{W}$. Two fouling
resistances add $2 \times 0.0005 = 0.001$, so $1/U_{\text{dirty}} =
0.00143 + 0.001 = 0.00243$ and $U_{\text{dirty}} = 412\,\si{\watt\per
\meter\squared\per\kelvin}$. At a fixed driving difference the duty
scales with $U$, so the capacity lost is $1 - 412/700 = 41\,\%$. The
fouling allowance the designer must build in to keep meeting the duty
at end-of-cycle is this same $41\,\%$ of extra clean capacity.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the LMTD expression by integrating the differential heat-
balance $-d(T_h - T_c)/(T_h - T_c) = U \,dA/C^*$ across the exchanger
length, where $1/C^* = 1/C_h \pm 1/C_c$ for counter-flow and parallel-
flow respectively.

\paragraph{Solution sketch.} Write $\theta = T_h - T_c$. The local heat
transfer across an element of area $dA$ is $d\dot{Q} = U\theta\,dA$,
and this heats the cold stream and cools the hot stream, so $dT_h = -d
\dot{Q}/C_h$ and $dT_c = \pm d\dot{Q}/C_c$ (the sign distinguishing
counter-flow from parallel-flow). Then $d\theta = dT_h - dT_c = -d
\dot{Q}(1/C_h \pm 1/C_c) = -U\theta\,dA/C^*$. Separating, $d\theta/
\theta = -(U/C^*)\,dA$, and integrating over the whole exchanger gives
$\ln(\theta_2/\theta_1) = -UA/C^*$. The endpoints fix $C^*$: since
$\theta_1 - \theta_2 = \dot{Q}/C^* \cdot (\text{constant})$, one
eliminates $C^*$ using $\dot{Q} = C^*(\theta_1 - \theta_2)$ and the
overall balance $\dot{Q} = UA\,\Delta T_{\text{LMTD}}$, which forces
$\Delta T_{\text{LMTD}} = (\theta_1 - \theta_2)/\ln(\theta_1/\theta_2)$.
The constant-$U$ and constant-$c_p$ assumptions are what make $C^*$
constant along the length.
\end{exercise}

\begin{exercise}
Derive the counter-flow effectiveness-NTU formula $\varepsilon = (1
- e^{-\text{NTU}(1 - C_r)})/(1 - C_r e^{-\text{NTU}(1 - C_r)})$.

\paragraph{Solution sketch.} Start from the integrated balance of the
previous exercise, $\theta_2/\theta_1 = e^{-UA(1/C_h - 1/C_c)} =
e^{-\text{NTU}(1 - C_r)}$ when $C_{\min} = C_h$ (so that $UA/C_h =
\text{NTU}$ and $C_h/C_c = C_r$). Call this ratio $X$. Write the two
end differences in terms of the effectiveness: $\theta_1 = T_{h,i} -
T_{c,o}$ and $\theta_2 = T_{h,o} - T_{c,i}$, and use $\dot{Q} =
\varepsilon C_{\min}(T_{h,i} - T_{c,i})$ to express $T_{h,o}$ and
$T_{c,o}$ as $T_{h,i} - \varepsilon(T_{h,i} - T_{c,i})$ and $T_{c,i} +
\varepsilon C_r(T_{h,i} - T_{c,i})$ respectively. Substituting into
$\theta_2 = X\theta_1$ and dividing through by $(T_{h,i} - T_{c,i})$
gives $1 - \varepsilon = X(1 - \varepsilon C_r)$. Solving for
$\varepsilon$ yields $\varepsilon = (1 - X)/(1 - C_r X)$, which is the
stated formula with $X = e^{-\text{NTU}(1 - C_r)}$. The $C_r = 1$ case
is the limit $X \to e^{0} = 1$, evaluated by l'Hopital's rule to give
$\varepsilon = \text{NTU}/(1 + \text{NTU})$.
\end{exercise}

\begin{exercise}
Show that for an evaporator or condenser, $\varepsilon = 1 -
e^{-\text{NTU}}$ regardless of the exchanger configuration.

\paragraph{Solution sketch.} A phase-changing stream holds constant
temperature, which is the limit $C_{\max} \to \infty$ on that side, so
$C_r = C_{\min}/C_{\max} \to 0$. Set $C_r = 0$ in any of the
configuration relations. Counter-flow: $\varepsilon = (1 - e^{-
\text{NTU}})/(1 - 0) = 1 - e^{-\text{NTU}}$. Parallel-flow: $(1 - e^{-
\text{NTU}})/(1 + 0) = 1 - e^{-\text{NTU}}$. Cross-flow: the exponent
$[\exp(-\text{NTU}^{0.78}C_r) - 1]\text{NTU}^{0.22}/C_r$ tends to
$-\text{NTU}$ as $C_r \to 0$ (since $(\exp(-aC_r) - 1)/C_r \to -a$),
giving $1 - e^{-\text{NTU}}$ again. All configurations collapse because
with one temperature fixed there is no upstream-versus-downstream
distinction for the driving difference, which is why condensers and
evaporators are easier to rate than sensible-heat exchangers.
\end{exercise}

\begin{exercise}
Derive the optimal flow velocity for a fixed heat-transfer area
(minimum of total cost = capital + pumping cost), assuming $h \propto
u^{0.8}$ and $\Delta p \propto u^2$.

\paragraph{Solution sketch.} For a fixed duty the required area scales as
$A \propto 1/h \propto u^{-0.8}$, so the capital cost is $C_{\text{cap}}
= k_1 u^{-0.8}$. The pumping power scales as $\dot{V}\,\Delta p \propto
u \cdot u^2 = u^3$, so the operating cost is $C_{\text{pump}} = k_2
u^{3}$. The total is $C(u) = k_1 u^{-0.8} + k_2 u^{3}$. Setting
$dC/du = -0.8 k_1 u^{-1.8} + 3 k_2 u^{2} = 0$ gives $u^{3.8} = 0.8
k_1/(3 k_2)$, so the optimum is $u^\star = [0.8 k_1/(3 k_2)]^{1/3.8}$.
The exponent $1/3.8 \approx 0.26$ makes the optimum insensitive to the
cost coefficients: the cost curve is flat near the minimum, so any
velocity within roughly a factor of two of $u^\star$ is near-optimal.
The routine \texttt{pressure\_drop\_optimum.py} confirms the analytic
result against a numerical scan.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the heat-transfer area required to condense $1\,\text{kg/s}$
of saturated steam at $40\,\degC$ using $25\,\degC$ cooling water with
a $10\,\si{\kelvin}$ temperature rise. State the assumed $U$.

\paragraph{Solution sketch.} The latent heat of steam near $40\,\degC$ is
about $2400\,\text{kJ/kg}$, so $\dot{Q} \approx 1 \times 2400 = 2.4\,
\text{MW}$. The water enters at $25\,\degC$ and leaves at $35\,\degC$,
so $\Delta T_1 = 40 - 35 = 5\,\si{\kelvin}$, $\Delta T_2 = 40 - 25 =
15\,\si{\kelvin}$, and $\Delta T_{\text{LMTD}} = (5 - 15)/\ln(5/15) =
9.1\,\si{\kelvin}$. Assume $U = 2500\,\si{\watt\per\meter\squared\per
\kelvin}$ for a steam-to-water condenser. Then $A = \dot{Q}/(U \Delta
T_{\text{LMTD}}) = 2.4 \times 10^6/(2500 \times 9.1) = 105\,\text{m}^2$.
The small approach at the cold end ($5\,\si{\kelvin}$) is what inflates
the area; condensers near the cooling-water inlet temperature are
area-hungry.
\end{exercise}

\begin{exercise}
Estimate the pumping power for the cooling-water side of a $1\,\text{MW}$
condenser, assuming $5\,\text{kg/s}$ of cooling water through tubes
of $25\,\si{\milli\meter}$ diameter at $2\,\text{m/s}$.

\paragraph{Solution sketch.} One tube at $2\,\text{m/s}$ carries $\rho u
A_{\text{tube}} = 990 \times 2 \times \pi/4 \times 0.025^2 = 0.97\,
\text{kg/s}$, so $5\,\text{kg/s}$ needs about five tubes (or a single
pass of five tubes). Take a tube length of $\sim 4\,\text{m}$. The
Reynolds number is $Re = \rho u D/\mu \approx 990 \times 2 \times
0.025/(8 \times 10^{-4}) \approx 6 \times 10^4$, giving a friction
factor $f \approx 0.02$. The pressure drop is $\Delta p = f(L/D)\rho
u^2/2 = 0.02 \times (4/0.025) \times 990 \times 2^2/2 \approx 6.3\,
\text{kPa}$. The volumetric flow is $\dot{V} = 5/990 = 5.0 \times
10^{-3}\,\text{m}^3/\text{s}$, so the hydraulic power is $\dot{V}\Delta
p \approx 32\,\text{W}$; with a pump efficiency of $0.6$ and adding
nozzle and header losses (roughly doubling the tube-only drop), the
shaft power is of order $100\,\text{W}$. Tube-side pumping for a
condenser of this size is small against the heat duty, which is the
usual finding for water service at sensible velocities.
\end{exercise}

\begin{exercise}
Estimate the volume of a hot-water thermal storage tank that holds
enough energy to heat a $200\,\text{m}^2$ house for $12\,\text{hours}$
in winter, with the tank cycling between $80\,\degC$ and $50\,\degC$.

\paragraph{Solution sketch.} A reasonably insulated house loses on the
order of $50\,\text{W/m}^2$ of floor area in cold weather, so a $200\,
\text{m}^2$ house has a heating load of about $10\,\text{kW}$. Over
$12\,\text{hours}$ the energy needed is $10 \times 12 = 120\,\text{kWh}
= 4.3 \times 10^8\,\text{J}$. A water tank cycling $\Delta T = 30\,
\si{\kelvin}$ stores $\rho c_p \Delta T$ per cubic metre $= 990 \times
4180 \times 30 = 1.24 \times 10^8\,\text{J/m}^3 \approx 34\,\text{kWh/m}^3$.
The tank volume is $120/34 \approx 3.5\,\text{m}^3$, a tank roughly
$1.5\,\text{m}$ on a side. The result scales directly with the assumed
heat-loss rate, so a leakier house or a colder night would push the
tank toward $5$ to $6\,\text{m}^3$.
\end{exercise}

\begin{exercise}
Estimate the surface area of a district-heating distribution pipe of
$200\,\si{\milli\meter}$ diameter and the heat loss per metre at
$80\,\degC$ internal, $5\,\degC$ external, with $100\,\si{\milli\meter}$
of polyurethane insulation.

\paragraph{Solution sketch.} The pipe outer radius is $r_1 = 0.1\,\text{m}$
and the insulation outer radius is $r_2 = 0.1 + 0.1 = 0.2\,\text{m}$.
The surface area per metre at the insulation surface is $2\pi r_2 =
1.26\,\text{m}^2/\text{m}$. Heat loss is dominated by conduction through
the insulation, $\dot{Q}/L = 2\pi k (T_{\text{in}} - T_{\text{out}})/
\ln(r_2/r_1)$ with polyurethane $k \approx 0.025\,\si{\watt\per\meter
\per\kelvin}$: $\dot{Q}/L = 2\pi \times 0.025 \times (80 - 5)/\ln(0.2/
0.1) = 11.8/0.693 = 17\,\text{W/m}$. The internal film and the pipe
wall add negligible resistance against the insulation, so the estimate
stands. Over a kilometre of supply main this is $17\,\text{kW}$ of
standing loss, which is why fourth- and fifth-generation networks push
the supply temperature down.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a one-dimensional finite-difference solver for a counter-
flow heat exchanger. Compare the simulated outlet temperatures to
the LMTD-method prediction for a test case.

\paragraph{Solution sketch.} Discretise the length into $N$ cells. March
the hot stream from its inlet end, $T_h^{i+1} = T_h^{i} - U\,\Delta A
(T_h^{i} - T_c^{i})/C_h$, and the cold stream from its inlet end (the
opposite end for counter-flow), $T_c^{i-1} = T_c^{i} + U\,\Delta A
(T_h^{i} - T_c^{i})/C_c$. Because the cold inlet boundary sits at the
far end, sweep the pair of marches repeatedly until the profiles stop
changing (a relaxation iteration), or assemble and solve the linear
system directly. The reference routine
\texttt{counterflow\_fd.py} implements the relaxation form and prints
both the finite-difference outlets and the closed-form effectiveness
prediction; for the supplied test case ($U = 1500$, $A = 3\,\text{m}^2$,
$C_h = 8400$, $C_c = 16{,}800$, hot in $80\,\degC$, cold in $20\,\degC$)
the two agree to better than $0.05\,\si{\kelvin}$, with the small
residual being the truncation error that shrinks as $N$ grows.
\end{exercise}

\begin{exercise}
Implement an iterative LMTD-based sizing routine: given the inlet
temperatures, mass flows, and one outlet temperature, compute the
other outlet temperature and the area required. Verify against a
textbook example.

\paragraph{Solution sketch.} The second outlet follows in closed form from
the energy balance: $\dot{Q} = C_{\text{known}}\,\Delta T_{\text{known}}$,
then the other stream's outlet from $\dot{Q} = C_{\text{other}}\,\Delta
T_{\text{other}}$. With all four temperatures in hand the LMTD is direct
and the area is $A = \dot{Q}/(U F \Delta T_{\text{LMTD}})$, so no
iteration is needed when both outlets reduce to balances. An iteration
is required only when $U$ depends on the (unknown) wall temperatures or
when the correction factor $F$ depends on terminal temperatures that
are themselves being solved; then loop: guess outlets, compute $F$ and
$U$, recompute the area and outlets, repeat to convergence. The routine
\texttt{lmtd\_sizing.py} performs the non-iterative core; verify it on
the $1\,\text{MW}$ steam-condenser example of the estimation block,
which it reproduces at $\Delta T_{\text{LMTD}} = 19.6\,\si{\kelvin}$ and
$A = 25.5\,\text{m}^2$.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify the geometry of a shell-and-tube preheater to recover heat
from a $5\,\text{kg/s}$ stream of hot exhaust gas ($200\,\degC$
inlet) to preheat $3\,\text{kg/s}$ of inlet air from $20\,\degC$.
Target a thermal effectiveness of $\varepsilon \geq 0.7$.

\paragraph{Solution sketch.} The capacity rates (air $c_p \approx 1005\,
\text{J/(kg K)}$) are $C_{\text{exhaust}} = 5 \times 1005 = 5025\,
\si{\watt\per\kelvin}$ and $C_{\text{air}} = 3 \times 1005 = 3015\,
\si{\watt\per\kelvin}$, so $C_{\min} = 3015$ (the inlet air), $C_r =
3015/5025 = 0.6$. For a counter-flow arrangement at $\varepsilon =
0.7$, invert the relation: $\text{NTU} = \frac{1}{C_r - 1}\ln\frac{
\varepsilon - 1}{\varepsilon C_r - 1} = \frac{1}{-0.4}\ln\frac{-0.3}{
-0.58} = \frac{1}{-0.4}\ln(0.517) = 1.65$. Then $UA = \text{NTU}\,
C_{\min} = 1.65 \times 3015 = 4975\,\text{W/K}$. Gas-to-gas
coefficients are low, $U \approx 25\,\si{\watt\per\meter\squared\per
\kelvin}$, so the area is $A = 4975/25 \approx 200\,\text{m}^2$, which
points to a finned cross-flow or compact recuperator rather than a bare
shell-and-tube. State the geometry: extended (finned) surface, a
counter-flow-biased arrangement to hold the inverted $F$, and tube
velocities chosen to keep the gas-side pressure drop within the
available draft.
\end{exercise}

\begin{exercise}
Design a domestic hot-water-to-radiant-floor distribution loop. The
floor area is $80\,\text{m}^2$ with $50\,\text{W/m}^2$ heat-loss
rate; the supply water is at $45\,\degC$ and the return water at
$35\,\degC$. Specify the loop spacing, pipe diameter, and flow rate.

\paragraph{Solution sketch.} The total heat delivered is $\dot{Q} = 80
\times 50 = 4\,\text{kW}$. The supply-to-return drop is $\Delta T =
45 - 35 = 10\,\si{\kelvin}$, so the water flow is $\dot{m} = \dot{Q}/
(c_p \Delta T) = 4000/(4180 \times 10) = 0.096\,\text{kg/s} \approx
0.1\,\text{kg/s}$, about $0.35\,\text{m}^3/\text{h}$. Radiant floor
practice spaces the tubing $150$ to $200\,\si{\milli\meter}$ on centre;
at $200\,\si{\milli\meter}$ spacing the $80\,\text{m}^2$ floor needs
roughly $80/0.2 = 400\,\text{m}$ of tube, which exceeds the length one
loop can carry at acceptable pressure drop, so split into about three
loops of $130\,\text{m}$ on a manifold. Use $16\,\si{\milli\meter}$
inner-diameter cross-linked polyethylene tube; at $0.1\,\text{kg/s}$
split three ways the per-loop velocity is modest ($\sim 0.2\,\text{m/s}$)
and the per-loop pressure drop stays within the circulator's head. The
binding constraints are the maximum floor surface temperature (comfort
and finish limits, near $29\,\degC$) and the per-loop length that keeps
the pressure drop pumpable.
\end{exercise}

\begin{exercise}
Specify the thermal-energy-storage tank for a concentrated solar
power plant: $100\,\text{MW}_e$ output, $6\,\text{hours}$ of storage,
molten-salt working fluid at $290$/$565\,\degC$. State the tank
volume, the heat-exchanger duty, and the temperature stratification
assumed.

\paragraph{Solution sketch.} Assume a power-block thermal efficiency of
about $0.4$, so $100\,\text{MW}_e$ for $6\,\text{hours}$ needs thermal
energy $E = (100/0.4) \times 6 = 1500\,\text{MWh}_{\text{th}} = 5.4
\times 10^{12}\,\text{J}$. Solar-salt stores $\rho c_p \Delta T = 1850
\times 1500 \times (565 - 290) = 7.6 \times 10^{8}\,\text{J/m}^3$. The
salt volume is $5.4 \times 10^{12}/7.6 \times 10^{8} = 7100\,\text{m}^3$.
Commercial plants use a two-tank arrangement (a cold tank at $290\,
\degC$ and a hot tank at $565\,\degC$), so two tanks of about $7000\,
\text{m}^3$ each, or roughly $40\,\text{m}$ diameter by $14\,\text{m}$
tall, rather than relying on stratification within a single
thermocline tank. The discharge heat-exchanger (salt-to-steam) duty is
the thermal rate, $100/0.4 = 250\,\text{MW}_{\text{th}}$. Two-tank
storage sidesteps the stratification question by physically separating
hot and cold inventories; a single-tank thermocline design would assume
a thermocline a metre or two thick and accept a few percent of usable
capacity lost to it.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A cooling-water-to-process-fluid heat exchanger has shown a
declining $U$ over the past year, from $1200$ to $700\,\text{W}/(
\text{m}^2\cdot\si{\kelvin})$. Identify the most probable fouling
mechanism and the diagnostic measurement.

\paragraph{Solution sketch.} A cooling-water service that loses $42\,\%$
of its coefficient over a year, on a roughly asymptotic decline, points
to crystallisation (scale) fouling on the water side, with biological
fouling as the second candidate if the water is warm and the velocity
low. The diagnostic is to track the coefficient and the pressure drop
together: scale raises the pressure drop modestly while it cuts $U$
sharply, whereas a step change in $U$ without a pressure-drop change
suggests an air or gas blanket rather than a deposit. Confirm by pulling
a tube at the next shutdown and identifying the deposit (calcium
carbonate scale versus biofilm versus silt), and check the water
chemistry (hardness, temperature, biocide residual). The dataset
\texttt{condenser\_U\_decline.csv} shows the matching coefficient and
pressure-drop signatures of a scaling service.
\end{exercise}

\begin{exercise}
A shell-and-tube exchanger in a refinery has begun to leak after
$15\,\text{years}$ of service. Identify the most probable corrosion
mode, the diagnostic, and the corrective action.

\paragraph{Solution sketch.} A leak appearing after long service, rather
than early, suggests a progressive wall-thinning or cracking mode
rather than a manufacturing defect. The leading candidates in refinery
service are erosion-corrosion at flow disturbances (inlet nozzles,
U-bends), under-deposit pitting beneath fouling, and, in chloride or
sulphide environments with stainless tubes, stress-corrosion cracking.
The diagnostic is eddy-current or, for ferromagnetic tubes,
remote-field testing of the full bundle to map wall loss and crack
indications, combined with locating the leaking tube by pressure-decay
or helium testing. The corrective action is to plug or sleeve the failed
tubes if few, retube if many, and address the root cause: a material
upgrade (for example to a duplex or titanium tube), an inlet-erosion
sleeve, or a fouling-control change to remove the under-deposit
mechanism.
\end{exercise}

\begin{exercise}
A district-heating substation's return temperature has risen from
$45\,\degC$ to $55\,\degC$ over a heating season. Identify what the
change indicates about the consumer-side hydraulic balance and the
diagnostic that distinguishes the possibilities.

\paragraph{Solution sketch.} A rising return temperature means the
substation is extracting less temperature drop from the network water
for the same heat, which is a fault, because high return temperature
raises distribution losses and limits the network's capacity. The
candidate causes are a fouled or undersized consumer heat exchanger
(too little area to cool the return), a control valve stuck open or
hunting (excess flow, short residence time), or a bypass left open that
returns hot supply water unused. Distinguish them by reading flow and
both temperatures at the substation: a fouled exchanger shows a falling
overall coefficient with normal flow; an open control valve or bypass
shows excess flow at low temperature drop. The fix follows the cause:
clean or enlarge the exchanger, re-tune or replace the valve, or close
the bypass. This is the consumer-side counterpart of the network
hydraulic design of section 6.6.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A nuclear-plant steam generator has been found to have a small
primary-to-secondary tube leak (detected by activity monitors). The
tube material is Inconel 600. Identify the most probable failure
mode (stress-corrosion cracking, denting, wastage) and state how the
detection method establishes which.

\paragraph{Solution sketch.} Inconel 600 in pressurised-water steam
generators is historically prone to primary-water stress-corrosion
cracking, especially at the tube-sheet expansion transitions and the
U-bend apexes where residual stress is highest, so cracking is the
leading mode over denting (caused by support-plate corrosion product
growth) and wastage (caused by phosphate water chemistry, now largely
designed out). The activity monitors establish only that a
primary-to-secondary leak exists, not its mechanism; the mechanism is
established by eddy-current inspection, which images crack-like
indications at the high-stress locations and distinguishes axial and
circumferential cracking from the volumetric metal loss of wastage and
the geometry change of denting. The corrective action is to plug or
sleeve the affected tubes and, for a fleet problem, to replace the
steam generator with thermally treated Inconel 690 tubing, which is far
more resistant to the cracking mode.
\end{exercise}

\begin{exercise}
A municipal cooling-tower system has suffered Legionnaires' disease
outbreaks among nearby residents. Identify the heat-transfer system
component responsible and the engineering controls that should have
prevented the outbreak.

\paragraph{Solution sketch.} The responsible component is the open
recirculating cooling tower, an evaporative direct-contact exchanger
whose warm water and drift aerosol are an ideal vector for
\emph{Legionella} bacteria carried on the drift droplets to nearby
people. The water sits in the temperature band ($25$ to $45\,\degC$)
that favours bacterial growth, the basin and packing offer biofilm and
nutrient surfaces, and the fan disperses fine droplets beyond the
tower. The engineering controls that should have been in place are a
maintained biocide dosing programme, regular cleaning and inspection of
the basin and fill to remove biofilm and sediment, high-efficiency
drift eliminators to cut the aerosol carried out of the tower, and a
water-management plan with periodic \emph{Legionella} sampling. The
case is described here by mechanism; no named registry entry is cited.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer argues that an exchanger ``with greater area is
always better.'' Write a one-paragraph reply that identifies the
costs of over-sized area (capital, pressure drop, fouling rate,
weight, installation).

\paragraph{Solution sketch.} A model reply: more area costs more metal,
more shell, and more support steel, so capital and installed weight
rise directly. If the extra area comes from more or longer tubes at the
same flow, the per-pass velocity may fall, which lowers the
heat-transfer coefficient and, worse, lets particulate and biological
fouling settle faster, so the over-sized unit can foul sooner and lose
the very margin it was bought for. Larger frontal area or more passes
also changes the pressure drop, raising pumping cost on one side or
forcing a velocity compromise. The right amount of area is the duty
plus a justified fouling allowance, not the largest unit that fits; the
junior engineer is optimising one variable while ignoring the four it
trades against.
\end{exercise}

\begin{exercise}
A facilities manager proposes to extend the cleaning interval on a
shell-and-tube cooling exchanger from six months to eighteen.
Write a one-paragraph reply that identifies the trade-off between
cleaning cost and degraded-performance cost and the data needed to
make the decision.

\paragraph{Solution sketch.} A model reply: extending the interval saves
direct cleaning cost (labour, the cost of taking the unit out of
service, chemicals) but lets the coefficient decline further along the
asymptotic-fouling curve, so the exchanger runs longer at degraded
duty, which costs lost throughput, higher pumping power against a rising
pressure drop, and a greater risk of under-deposit corrosion that
shortens tube life. The decision needs the fouling-decline data for
this unit (the coefficient and pressure-drop history that fixes how
fast performance falls), the marginal value of the lost duty, the
cleaning cost and outage duration, and the corrosion risk of leaving
deposits in place longer. The honest answer is to set the interval
where the marginal degraded-performance cost over the next month equals
the amortised cleaning cost, which the unit's own decline curve and the
plant's value of throughput determine; eighteen months may be right or
reckless depending on those numbers.
\end{exercise}
