\chapter{Manufacturability, maintainability, sustainability, accessibility}
\label{vol11:ch10}

\epigraph{The design's complete life is its substance; designs that succeed in manufacture and fail in service, or that succeed in service and fail at end of life, have under-served their lifecycle.}{}

\chapmeta{Half-life: medium-life. Archetypes: ethics, scaling. Exercise target: 20. Project track: hybrid.}

\noindent
The four "ilities" of this chapter, manufacturability, maintainability, sustainability, and accessibility, are the lifecycle disciplines that complement the functional and performance design of earlier chapters. A design that meets its functional requirements but cannot be manufactured at acceptable cost is not a design; a design that meets its functional requirements but cannot be maintained in service is not a design; a design that meets its functional requirements but produces environmental damage that exceeds its benefit is not a design; a design that meets its functional requirements but excludes substantial fractions of its intended user population is not a design. This chapter develops the four lifecycle disciplines as integrated design considerations rather than as end-of-process add-ons.

\section{Design for manufacture}\pathtag{core}

Design for manufacture (DFM) is the discipline of designing artefacts that can be produced at acceptable cost, quality, and rate using the available manufacturing processes. The discipline is partly the design's compatibility with the process (the part can be machined, cast, moulded, additively manufactured), partly the design's economics under the process (the cost per unit at the target volume), and partly the design's quality in production (the process consistency produces conforming units at acceptable yield).

The DFM principles, abstracted across manufacturing processes:

\textbf{Match process to design.} Different processes suit different geometries, materials, volumes, and tolerances. Injection moulding is appropriate for high-volume plastic parts with moderate tolerances; CNC machining is appropriate for moderate-volume metal parts with tight tolerances; additive manufacturing is appropriate for low-volume parts with complex geometries; investment casting is appropriate for moderate-volume parts with complex internal features. Mismatching produces designs that the chosen process cannot economically produce.

\textbf{Design within process capability.} The chosen process has limits on geometry (minimum wall thickness; maximum aspect ratio; maximum part size), tolerances (positional accuracy; surface finish), and materials (which materials the process can handle). Designs that exceed the process capability require non-standard processes or process modifications, both of which raise cost.

\textbf{Minimise part count.} Each part adds manufacturing cost (the part's own production plus its assembly into the larger product). The DFM discipline includes consolidating multiple parts into single parts where the consolidation is feasible (single moulded piece versus assembly of multiple pieces; integral features versus separately-attached pieces). The classical example is the integral hinge in plastic packaging: a single moulded part with a thinned region that flexes, replacing a separate hinge with fasteners.

\textbf{Standardise.} The design uses standard components, standard materials, standard processes wherever the function permits. Standardisation reduces cost (standard items have economies of scale; standard processes have established capability), reduces lead time (standard items are available; non-standard items require sourcing), and reduces risk (standard items have established quality).

\textbf{Avoid manufacturing-hostile features.} Some features are characteristically difficult to manufacture: tight tolerances on hard-to-access surfaces; sharp internal corners (require special tooling); features that require multiple setups on a machining centre; very thin walls in injection-moulded parts (require careful gating and cooling); deep holes with high aspect ratios (require special drilling techniques). The DFM discipline includes redesigning to avoid these features where the function permits.

\textbf{Tolerance allocation.} The tolerance budgets of Chapter 5 are partly DFM decisions: tolerances should be no tighter than the function requires; tighter tolerances increase manufacturing cost; the cost-effective tolerance allocation matches function need to process capability.

\textbf{Design for test.} Manufacturing produces units that must be tested for conformance. The design includes provisions for test (test points; access for measurement; built-in self-test for electronics); designs without test provisions require destructive or expensive test methods.

\textbf{Process documentation.} The DFM work produces the documentation the manufacturer uses: process specifications; work instructions; quality plans; inspection procedures. The documentation is part of the design deliverable; designs without manufacturing documentation are incomplete.

The DFM discipline is taught in mechanical-engineering programmes (Boothroyd, Dewhurst, and Knight's "Product Design for Manufacture and Assembly" is a foundational reference) and is codified in industry practices (the IPC standards for electronics; the SAE standards for automotive; the various industry-specific practices). Modern CAD tools include DFM analysis features that flag potential manufacturing issues during detailed design; the engineer's discipline is to address the flags rather than to ignore them.

\begin{principle}[Design for what gets built]
A design that cannot be manufactured at acceptable cost is not a design; it is a specification of an artefact that will not exist. The DFM discipline integrates manufacturing knowledge into design decisions from early stages, not as a downstream review. Designs that engage manufacturing late are reliably more expensive to produce and more difficult to redesign than designs that engage manufacturing early.
\end{principle}

\section{Design for assembly}\pathtag{standard}

Design for assembly (DFA) is the DFM subdiscipline focused on the assembly process: how the manufactured parts are joined into the final product. Assembly is often a substantial fraction of total manufacturing cost (especially for products with many parts); DFA-aware design can reduce the cost substantially.

The DFA principles:

\textbf{Minimise part count.} Each part requires handling, orienting, and joining; reducing the part count reduces the assembly effort. The DFM minimisation overlaps with the DFA minimisation; they are not always identical (a single complex part may be cheaper to manufacture as multiple simpler parts but more expensive to assemble).

\textbf{Self-locating features.} Parts that self-locate during assembly (geometric features that constrain the assembly position) reduce the operator's positioning work. Examples: chamfers that guide a pin into a hole; locating pins that establish position before fastening; clamshell features that mate uniquely in one orientation.

\textbf{One-direction assembly.} Assemblies that are built up in a single direction (typically vertically, with gravity assisting the alignment) are easier and more consistent than assemblies that require multiple directional moves. The principle dominates consumer-electronics assembly: each component is placed from above onto the PCB; the PCB is mounted into the case from above; the case is closed from above.

\textbf{Symmetric parts.} Parts that are symmetric in the assembly direction can be installed without operator orientation; asymmetric parts require orientation that adds time and creates error opportunities. The DFA discipline either makes parts symmetric where the function permits or makes parts so obviously asymmetric that mis-orientation is impossible.

\textbf{Combine fasteners.} Each fastener adds assembly time; designs that use fewer fasteners (combined fasteners; snap fits; integral retention) reduce the assembly time. The trade-off is serviceability: fewer fasteners may make disassembly for service more difficult.

\textbf{Eliminate adjustments.} Assemblies that require post-assembly adjustment (alignment; calibration; tuning) require additional labour and create variation; designs that achieve the necessary alignment through tolerance discipline at part level eliminate the adjustment step.

\textbf{Layer-by-layer assembly.} Assemblies organised in layers (PCB; housing layers; covers) are easier to disassemble for rework or service than assemblies organised in nested geometries that require disassembly of multiple layers to access internal parts.

The Boothroyd-Dewhurst DFA methodology provides a quantitative DFA index: parts are scored on their handling and insertion difficulty; the design's total assembly time is computed; the design is compared to the theoretical minimum (assembly with only the functionally-necessary parts). The methodology is widely used in industry; it produces concrete redesign suggestions.

\section{Design for maintenance}\pathtag{core}

Design for maintenance (DFMt) is the discipline of designing artefacts that can be maintained in service: serviced; repaired; calibrated; upgraded; eventually replaced. The discipline is the response to the lifecycle reality that no artefact operates in service without maintenance, and that maintenance cost over the artefact's lifecycle often exceeds the original purchase cost.

The DFMt principles:

\textbf{Access to serviceable items.} The items that require periodic service (filters; lubrication points; consumable batteries; sensors that drift) are accessible without major disassembly. The accessibility may be through service panels, removable covers, or design layouts that place service items at the periphery.

\textbf{Modular replacement.} The artefact is composed of modules that can be replaced individually rather than requiring whole-product replacement. Modern automotive electronics, aerospace LRUs (line-replaceable units), and field-replaceable modules in IT infrastructure are dominant examples.

\textbf{Diagnostics.} The artefact's failure modes are detectable: built-in self-test; status indicators; diagnostic interfaces; logged events. The diagnostics allow service personnel to identify the failed item rapidly rather than through trial-and-error replacement.

\textbf{Standard tools.} The artefact can be serviced with standard tools (standard wrench sizes; standard screwdriver bits; standard test instruments) rather than requiring proprietary tools. The standardisation reduces the service infrastructure required.

\textbf{Documentation.} Service manuals, parts catalogues, troubleshooting guides, and training materials are part of the design's deliverable. Designs without service documentation are designs that the service organisation cannot support.

\textbf{Parts availability.} The components used in the design are available for service over the artefact's planned service life. Discontinued components produce service difficulties; the DFMt discipline includes considerations of component lifecycle in selection.

\textbf{Failure mode informational design.} The artefact's failure modes are designed to be evident: a fault that the user can detect is preferable to a silent fault that the user cannot. The "fail-safe" design principle includes the "fail-evident" extension.

The cost of DFMt is real: design effort; potentially additional material for serviceability features (access panels; modular structures); design constraints that limit functional optimisation. The benefit is the lifecycle maintenance cost reduction and the customer's lifecycle value.

The repairability movement (the right-to-repair legislation in various jurisdictions; the iFixit-style repairability scoring) is pressing the design community to take DFMt seriously for consumer products. The current pattern in consumer electronics is anti-repairability: parts glued together; non-standard fasteners; software locks that disable repaired parts. The pattern is sometimes commercially motivated (replacement sales exceed repair revenue) but is regulatorily contested.

\section{Design for inspection}\pathtag{standard}

Design for inspection is the DFMt subdiscipline focused on the ongoing assessment of the artefact's condition in service: non-destructive testing access; instrumentation provisions; visual inspection access; performance-monitoring data collection.

The principles:

\textbf{NDT access.} Critical structural and welded joints are positioned where non-destructive testing (ultrasonic; radiographic; magnetic-particle; dye-penetrant; eddy-current) can be performed. NDT access often requires designed-in openings, removable covers, or geometrical features that admit the probe.

\textbf{Visual inspection.} Parts that benefit from visual inspection (for corrosion; for cracking; for fluid leaks) are visible without disassembly. The design may include inspection windows; reflective surfaces; light-coloured surfaces that reveal contamination.

\textbf{Instrumentation provisions.} The artefact is designed to accept instrumentation: sensors for monitoring; data interfaces for downloading; sample ports for fluid analysis. The instrumentation may be permanent (built into the design) or temporary (connected for periodic monitoring).

\textbf{Performance data collection.} The artefact's performance data are recorded during service: operating hours; load history; environmental exposure; events. The data support condition assessment (the digital-twin practice of Chapter 6) and forensic analysis if failures occur.

\textbf{Inspection scheduling.} The design includes recommended inspection intervals; the intervals are based on the failure-mechanism analysis (Volume X) and the design's specific characteristics. Service manuals capture the intervals.

The aviation industry's structural inspection programmes are the most mature: every aircraft component has a documented inspection regime; the regimes are updated based on fleet experience; the inspections produce the data that establish the airworthiness of the in-service fleet. The discipline is partly transferable to other domains (civil infrastructure; nuclear plants; industrial equipment) and is partly specific to aviation's unique combination of high stakes, regulatory framework, and operational discipline.

\section{Design for end-of-life}\pathtag{standard}

Design for end-of-life addresses what happens to the artefact when it is no longer in service: disposal; recycling; reuse; remanufacture. The discipline is the design's recognition that artefacts have lifecycles, and that the post-service disposition is part of the lifecycle.

The end-of-life pathways:

\textbf{Disposal to landfill or incineration.} The dominant historical pathway; the lowest-cost option in the short term; the most environmentally damaging in the long term. The pathway's appropriateness depends on the material's environmental persistence and toxicity.

\textbf{Recycling.} The artefact's materials are recovered and reused as inputs to new manufacturing. Recycling is well-established for some materials (aluminium; steel; glass; paper) and underdeveloped for others (mixed plastics; composites; electronics). The recyclability of the artefact depends on its material composition, its construction, and the infrastructure available for processing.

\textbf{Reuse.} The artefact (or its components) is reused for the same or a different function. Reuse is dominant in industrial equipment (rebuilt machinery), in consumer products (used cars; used electronics), and in some construction (reclaimed materials). The reuse pathway is environmentally preferable to recycling for many applications because it preserves the embodied energy.

\textbf{Remanufacturing.} The artefact is disassembled, components are refurbished, and the components are reassembled into a remanufactured product. The remanufacturing industry is substantial in some sectors (automotive parts; office equipment; certain industrial equipment); it produces near-equivalent products at substantially reduced material and energy cost.

\textbf{Energy recovery.} Materials that cannot be recycled are incinerated for energy. The pathway is preferable to landfill for combustible materials but produces emissions that may be regulated.

The design considerations:

\textbf{Material selection for recyclability.} Materials that are recyclable in the available infrastructure (commodity metals; major plastics) are preferred to materials that are not (composites; mixed materials; rare alloys).

\textbf{Disassembly design.} The artefact is designed to be disassembled at end of life: mechanical fasteners rather than adhesives; consistent material identification; minimal hazardous content. Disassembly design overlaps with DFA but with different priorities (the assembly direction may differ from the disassembly direction).

\textbf{Material identification.} The materials in the artefact are marked or coded so the recycler can identify and separate them. The major plastic recycling codes (PETE 1; HDPE 2; etc.) are an example.

\textbf{Hazardous content minimisation.} Materials with end-of-life environmental concerns (heavy metals; persistent organic pollutants; ozone-depleting refrigerants) are minimised or eliminated in favour of less-harmful alternatives. The EU's RoHS directive (Restriction of Hazardous Substances) and REACH regulation drive substantial design changes in electronics and chemical-content management.

\textbf{Extended producer responsibility.} Some jurisdictions require the producer to take back the artefact at end of life; the requirement creates economic incentive for designs that are economical to process at end of life.

The end-of-life pathway is partly designed and partly imposed by the regulatory and economic context. The design's role is to enable the pathway the lifecycle analysis identifies as preferable; the regulatory and economic context determines whether the enabled pathway is actually used.

\section{Sustainable design}\pathtag{core}

Sustainable design integrates environmental footprint considerations (Volume X Chapter 15) into the design's lifecycle. The discipline goes beyond end-of-life to address the design's full environmental impact: raw materials; manufacturing energy and emissions; use-phase consumption; service-phase impact; end-of-life.

The principles:

\textbf{Energy efficiency in use.} For products whose use-phase energy consumption is significant (vehicles; appliances; lighting; computing infrastructure), the use-phase impact often dominates the lifecycle impact. Design for efficiency in the use phase is the dominant lever for these products. The principles include: appropriate sizing for the actual demand; efficient components; energy-management software; minimisation of standby losses.

\textbf{Material efficiency.} The design uses materials efficiently: minimum mass consistent with the function (lightweight design); minimum waste in manufacturing (material-efficient processes; near-net-shape manufacturing); use of recycled-content materials where the function permits; use of bio-based materials where the function permits.

\textbf{Low-impact materials.} The materials selected have lower environmental impact than alternatives that would serve the same function: lower embodied carbon; lower water footprint; lower toxicity; lower extraction impact. The selection is informed by life-cycle assessment data (Section 7).

\textbf{Durability.} The design lasts: long service life; resistance to wear; protection against degradation. Durable designs amortise their embodied environmental impact over more service hours; throwaway designs concentrate the impact per service hour.

\textbf{Repairability and serviceability.} The DFMt principles support sustainability: artefacts that can be repaired serve their function longer; artefacts that cannot be repaired are replaced when single failures occur, with the full new-product environmental impact.

\textbf{End-of-life pathways.} The previous section's pathways are integrated into the design's lifecycle planning.

\textbf{System-level efficiency.} Some sustainability improvements come from system-level rethinking rather than from component optimisation: replacing private vehicles with shared mobility; replacing physical meetings with video conferencing; replacing physical products with digital alternatives. The system-level moves often exceed the component-level efficiency gains in total impact.

The discipline of sustainable design is integrated with life-cycle analysis (next section); the analytical methods produce the evidence that informs the design choices. The discipline is also integrated with the broader Volume X Chapter 15 treatment of planetary footprint: the engineer's responsibility for the design's footprint is the engineering response to the planetary-scale problem.

\section{Life-cycle analysis}\pathtag{standard}

Life-cycle analysis (LCA) is the methodology for quantifying the environmental impact of a product or process across its lifecycle. The methodology was developed in the 1990s and is codified in ISO 14040 (principles and framework) and ISO 14044 (requirements and guidelines).

The LCA methodology has four stages:

\textbf{Goal and scope definition.} The analyst defines what is being analysed: the product or process; the functional unit (the unit of service the analysis is based on; e.g., "1 km of passenger transport" for a vehicle; "1 kWh of energy delivered" for a power plant); the system boundary (cradle-to-grave; cradle-to-gate; cradle-to-cradle); the impact categories of interest.

\textbf{Inventory analysis.} The analyst compiles the energy and material flows across the system boundary: inputs (raw materials; energy; water; land); outputs (products; co-products; emissions to air, water, soil; waste). The inventory is the quantitative basis for the impact assessment.

\textbf{Impact assessment.} The inventory is translated into impact categories: climate change (CO$_2$-equivalent emissions); acidification (SO$_2$-equivalent); eutrophication (NO$_3$-equivalent); human toxicity; ecotoxicity; resource depletion; water use; land use; ozone depletion; particulate matter formation. The assessment uses characterisation factors that convert inventory flows into impact-category indicators.

\textbf{Interpretation.} The analyst draws conclusions: identification of the dominant impact categories; the dominant lifecycle stages; the dominant materials or processes. The interpretation informs the design decisions: where the impact reductions would matter most; where the design choices are most consequential.

The major LCA databases (ecoinvent; GaBi; USLCI; the agri-environmental databases) provide the characterisation factors and the standard inventory data for common materials and processes. The major LCA software tools (SimaPro; GaBi; OpenLCA; the various academic tools) implement the methodology and integrate the databases.

The discipline of using LCA in design:

\textbf{Streamlined LCA at concept stage.} At conceptual design, LCA is typically streamlined: focus on the dominant impact categories; use database values for materials and processes; produce comparative rather than absolute assessment. The streamlined approach informs concept selection without consuming the effort of full LCA.

\textbf{Detailed LCA at design freeze.} At design freeze, a more detailed LCA documents the design's expected lifecycle impact for stakeholder communication and for regulatory submission.

\textbf{Sensitivity analysis.} The LCA's sensitivity to data quality, methodological choices, and modelling assumptions is documented; the conclusions are reported with appropriate caveats about the analysis's confidence.

\textbf{Updating over the design's lifecycle.} The LCA is updated as the design matures, as the upstream supply chain changes, as the use-phase context changes, as the end-of-life infrastructure evolves. The LCA's currency supports its continued usefulness.

LCA's limitations are real: data quality varies; methodological choices affect conclusions; impact categories cannot be summed into a single score without subjective weighting; the analysis covers what is quantifiable and may miss what is not. The discipline of using LCA is the discipline of accepting these limitations and reporting the analysis with appropriate caveats.

\section{Accessibility as engineering practice}\pathtag{core}

Accessibility is the design discipline that ensures the artefact serves the widest plausible range of users, including users with disabilities and users in contexts that the dominant user population does not represent. The discipline is treated as engineering practice rather than as a compliance exercise, following the editorial decision settled for this volume (Q47).

The user populations that accessibility design addresses:

\textbf{Users with visual variation.} Low vision; blindness; colour-vision differences; light sensitivity. The design supports these users through: text and symbols sized appropriately; sufficient contrast between text and background; alternatives to colour-only information; screen-reader compatibility for software; tactile features for non-software products.

\textbf{Users with hearing variation.} Hearing loss; deafness; auditory processing differences. The design supports these users through: visual alternatives to auditory alerts; captioning for spoken content; vibration as an alternative signalling modality; compatible operation with hearing aids and cochlear implants.

\textbf{Users with motor variation.} Limited dexterity; tremor; paralysis; missing limbs; arthritis; strength limitations. The design supports these users through: large activation targets; minimal force requirements; alternative input modalities (voice; switch; eye tracking); freedom from required two-handed operation; compatibility with assistive technologies.

\textbf{Users with cognitive variation.} Learning differences; memory variation; processing-speed differences; neurodevelopmental variation. The design supports these users through: clear language; consistent layouts; predictable interaction patterns; recovery paths from errors; minimisation of working-memory demands.

\textbf{Users with speech variation.} Limited speech; speech differences; non-native language speakers. The design supports these users through: alternatives to voice-only interfaces; text-based interaction modes; appropriate language complexity.

\textbf{Situational and temporary limitations.} Users without disability but in situations that produce equivalent limitations: bright sunlight (equivalent to low vision); noisy environment (equivalent to hearing loss); carrying a child (equivalent to one-handed operation); injury (temporary motor limitation); fatigue (cognitive limitation). The accessibility framework treats the situational equivalents as part of the population the design must serve.

\textbf{Low-resource contexts.} Users in contexts with limited bandwidth, intermittent power, limited literacy, limited disposable income. The design supports these users through: low-bandwidth modes; offline operation; multi-modal interfaces; affordability.

The principles of accessibility design:

\textbf{Universal design.} The design serves the widest plausible range of users without requiring adaptation. Universal design is the aspiration; perfect universality is rarely achievable; the discipline is to approach universality and to provide explicit alternatives where universality is not feasible.

\textbf{Inclusive design.} The design is developed with input from the range of users it will serve; the input drives design decisions; the design's evaluation includes the range. The discipline overlaps with user research (Chapter 3) and brings the accessibility-relevant user populations into the user-research samples.

\textbf{Assistive-technology compatibility.} The design works with the assistive technologies users employ (screen readers; alternative keyboards; switch controls; magnification; voice control). The compatibility is verified through testing with the actual assistive technologies; the design's compatibility is part of its specification.

\textbf{Standards compliance.} The design meets the applicable accessibility standards: WCAG (Web Content Accessibility Guidelines) for web and software; Section 508 for US federal procurement; the EN 301 549 for European public procurement; ISO 9241 for ergonomic accessibility; the various national equivalents. Standards compliance is necessary but not sufficient; designs can be compliant in form and fall short in substance.

\textbf{Persona spectrum.} The design's user personas include the range of users; the design decisions are reviewed against the personas; the gaps that the personas surface are addressed.

The discipline of accessibility-as-engineering-practice treats accessibility as a substantive design input rather than as a compliance review at the end of design. The treatment produces designs that serve more users at less marginal cost than the compliance-driven alternative; it also produces designs whose accessibility features benefit users beyond those with the specific disabilities the features were designed for.

\section{Failure: parts that could not be repaired, replaced, or recycled}\pathtag{core}

The failure mode this section addresses is the design that cannot be serviced at end of life: parts that cannot be repaired (cracked, broken, worn out, with no replacement available); designs that cannot be recycled (mixed materials inseparable; hazardous content; bonded constructions); products that cannot be disassembled (glued together; non-standard fasteners; software locks).

The recurring patterns:

\textbf{Glued construction.} Consumer products (smartphones; tablets; some appliances) are increasingly assembled with adhesives rather than fasteners. The adhesive bond is unobtrusive in the assembled product but prevents non-destructive disassembly. The repairability and recyclability are both severely compromised.

\textbf{Discontinued components.} The design uses components whose lifecycle is shorter than the product's. When the component is discontinued, no replacement is available; the product's service life is bounded by the component's availability.

\textbf{Proprietary parts.} The design uses parts that are available only from the original manufacturer; the parts are priced as captive components. Third-party suppliers cannot enter the market; the price exceeds what a competitive market would set.

\textbf{Software locks.} The design includes software that detects replacement parts and refuses to operate; the lock prevents independent repair. The pattern is increasingly common in automotive (sensors locked to specific vehicles), in agricultural equipment (controllers locked to specific dealerships), in consumer electronics (display modules locked to specific phones).

\textbf{Inseparable materials.} The design combines materials in ways the recycling infrastructure cannot separate: composite plastics fused to metal; coatings that contaminate the substrate; mixed plastics with no marking. The product's end-of-life pathway is landfill regardless of regulatory intent.

\textbf{Hazardous content.} The design includes materials whose disposal is regulated: heavy metals; persistent organic pollutants; battery cells with specific handling requirements. The end-of-life processing requires specialised facilities; absent the facilities, the materials are mishandled.

\textbf{Undocumented configurations.} The design's configuration (firmware versions; software dependencies; calibration constants) is not documented in a form that subsequent service can recover. After the design's manufacturer exits the market, the surviving products cannot be supported.

The recent regulatory responses include:

\textbf{Right-to-repair legislation.} Multiple jurisdictions (EU; several US states; Canada) have legislated or are legislating requirements that manufacturers provide repair documentation, parts, and tools to independent repair providers.

\textbf{Extended producer responsibility.} The producer is required to manage the product's end-of-life; the regulatory framework creates economic incentive for designs that are economical at end of life.

\textbf{Eco-design directives.} The EU's Eco-design Directive and related instruments require energy efficiency, repairability, and recyclability criteria for many product categories.

\textbf{Material restrictions.} RoHS (restriction of hazardous substances); REACH (registration, evaluation, authorisation of chemicals); the various national equivalents restrict the materials that can be used.

The engineering response is the design discipline this chapter develops: integration of manufacturability, maintainability, sustainability, and accessibility considerations into design from early stages. The cost of the discipline is real (design effort; design constraints; cost increases in some cases); the cost of not exercising the discipline is the lifecycle failure mode this section addresses and the regulatory consequences the response provokes.

\begin{principle}[The design serves its full lifecycle]
A design that serves only the manufacturing phase produces products that fail in service; a design that serves only the use phase produces products that cannot be maintained; a design that serves only the new phase produces products that cannot be recycled; a design that serves only the dominant user produces products that exclude substantial populations. The discipline of design is the discipline of serving the artefact's full lifecycle and its full user population.
\end{principle}

\begin{archetype}[scaling]
The four lifecycle disciplines are the scaling archetype applied to the design's temporal extent. The design's performance over its lifecycle, not at any single moment, is what determines its actual value. The discipline of integrating manufacturability, maintainability, sustainability, and accessibility from early design stages produces designs whose lifecycle performance exceeds designs that integrate these considerations late or not at all.
\end{archetype}

\begin{project}[End-of-life redesign]
Select one previous-chapter project (the reader's detailed-design pack from Chapter 5; the prototype from Chapter 7; the test plan's product from Chapter 8). Redesign the product for end-of-life: identify the current design's end-of-life pathway; identify the design changes that would improve the pathway (disassembly; material selection; component reuse; recyclability); estimate the trade-offs (material cost; manufacturing complexity; performance impact); produce a redesigned version with the changes. Deliverable: a 3-page document with the original design's end-of-life characterisation, the proposed changes, the trade-off analysis, and the redesigned version.

\textbf{Hazard.} The exercise often reveals that the original design was developed without end-of-life consideration; the redesign reveals what would have been required to integrate the consideration from the start. The lesson is the engineering discipline of integration rather than the specific redesigned artefact.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A product has 50 parts in the original design; DFA analysis suggests the part count can be reduced to 25 through consolidation. The per-part assembly time is 30 seconds; the per-part cost is \$0.50 plus a \$0.25 inventory-management cost. Compute the assembly-time savings and the cost savings from the part-count reduction.
\end{exercise}

\begin{exercise}
A vehicle has an operational lifetime of 200{,}000 km. Its use-phase CO$_2$ emissions are 150 g/km; its manufacturing CO$_2$ is 8 tonnes; its end-of-life impact (after recycling credits) is $-1$ tonne. Compute the total lifecycle CO$_2$. Identify the dominant phase.
\end{exercise}

\begin{exercise}
A consumer product has a 4-year service life if not repaired; with a single repair (cost \$80), service life extends to 7 years. New-product cost is \$300. Compute the cost-per-year for repair-no, repair-yes; identify the consumer-economic argument for the repair option. Identify the lifecycle-CO$_2$ implication.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the optimal trade-off between part-count consolidation (lower assembly cost; higher per-part complexity) and modular replacement (higher assembly cost; lower service cost) for a product whose service-life cost depends on both. Identify the design parameters that shift the optimum.
\end{exercise}

\begin{exercise}
For a product's recycling decision, derive the conditions under which recycling is preferable to virgin material use, considering: collection cost; processing cost; transportation; quality degradation; market value of recycled material. Identify the conditions under which the recycling pathway is economically viable without subsidy and those under which it requires regulatory support.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the fraction of total manufacturing cost devoted to assembly for: a smartphone; an automobile; a commercial aircraft; a building. Identify the implications for DFA priority across the categories.
\end{exercise}

\begin{exercise}
Estimate the lifecycle CO$_2$ footprint of a typical laptop computer over its 4-year service life. Identify the dominant phase. Compare to a longer service life (6 or 8 years) achieved through repair and upgrade.
\end{exercise}

\begin{exercise}
Estimate the fraction of consumer electronic devices that reach end of life through landfill versus recycling versus reuse versus refurbishment. Discuss the design and policy interventions that could shift the distribution.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Implement a streamlined LCA of a simple product (e.g., a beverage bottle: glass vs. PET vs. aluminium). Use database values for material extraction, manufacturing, transportation, and end-of-life; compute the CO$_2$ footprint per functional unit; compare the three alternatives.
\end{exercise}

\begin{exercise}
Simulate the consumer-product repair decision over a population of devices with varying failure modes: each failure has a repair cost and a probability of recurring; the consumer's decision rule depends on the cost ratio. Sweep parameters and identify the conditions under which repair extends service life substantially versus marginally.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a product (a small consumer device) with explicit DFMt principles: accessibility of serviceable parts; modular construction; standard fasteners; diagnostics; documentation. Identify how the DFMt design differs from an equivalent design without DFMt consideration.
\end{exercise}

\begin{exercise}
Design an end-of-life pathway for a hypothetical electronic product: the manufacturer's take-back programme; the processing facility's operations; the material recovery; the residual disposition. Specify the design choices that support the pathway.
\end{exercise}

\begin{exercise}
Design an accessibility evaluation protocol for a software product. Specify the user populations represented, the assistive technologies tested, the WCAG requirements verified, the criteria for the design to pass.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: the current trend toward glued, unrepairable consumer products is a structural failure of the design profession that regulatory intervention (right-to-repair legislation; extended producer responsibility) is correctly addressing. Argue the opposite. Identify the empirical evidence that bears on the dispute.
\end{exercise}

\begin{exercise}
Discuss the role of accessibility in design. Argue: accessibility integrated from the start produces designs that serve more users at lower marginal cost than accessibility addressed as compliance. Identify the institutional and educational changes that would integrate accessibility into the engineering profession's standard practice.
\end{exercise}

\begin{exercise}
A product manager asks the design team to remove accessibility features from a product to reduce cost. The features are not legally required for the product's market. Discuss the design team's options under the engineering codes of ethics and the discipline of design developed in this chapter.
\end{exercise}

\begin{exercise}
Argue: sustainability considerations should be integrated into design as a first-order requirement equivalent to cost, schedule, and performance. Identify the structural conditions under which this integration is feasible and the structural conditions that resist it. Discuss the engineer's role in advocating for the integration.
\end{exercise}

\begin{exercise}
Discuss the right-to-repair regulatory movement. Identify what the design profession can do to support the movement substantively versus the actions that would be compliance-theatre. Argue that designs that anticipate right-to-repair requirements produce better designs even in jurisdictions where the requirements are not yet binding.
\end{exercise}

\begin{exercise}
Take one piece of equipment in the reader's home or workplace whose maintenance regime is documented. Compare the actual maintenance the equipment receives to the documented regime. Identify the gaps and the failure modes the gaps would produce. Identify whether the gaps are attributable to the equipment's design (DFMt failures) or to the maintenance organisation (operational failures).
\end{exercise}

\begin{exercise}
A design organisation operates without explicit life-cycle assessment in its design process. Discuss the practical steps to introduce streamlined LCA at concept stage: the personnel; the tools; the integration with concept selection; the cultural changes; the metrics that would evaluate the introduction's success.
\end{exercise}
