\chapter{Hydraulics, pneumatics, mechatronic integration}
\label{vol08:ch14}

\epigraph{The full machine works only when every subsystem works, and only when the interfaces between them are correct.}{}

\chapmeta{Half-life: medium-life. Archetypes: balance, interface, scaling. Exercise target: 25. Project track: physical.}

\noindent
A modern machine integrates mechanical, electrical, and digital subsystems into a coherent whole. The mechanical structure provides the geometry, the bearings, the linkages; the electrical subsystem provides actuation, sensing, and power; the digital subsystem provides control, communication, and intelligence; and increasingly, fluid power (hydraulics and pneumatics) provides force and motion where electric actuation is impractical. This final chapter of Volume VIII addresses two areas that complete the engineer's machine toolbox: fluid power as a complementary actuation technology, and mechatronic integration as the discipline that makes the full machine work. The reader who finishes this chapter has the working vocabulary to specify, design, integrate, and commission a complete engineering system.

\section{Fluid power fundamentals}\pathtag{standard}

Fluid power transmits force and motion through a pressurised fluid. The two families are hydraulics (essentially incompressible oil, typically at $50$ to $350\,\text{bar}$) and pneumatics (compressed air, typically at $6$ to $10\,\text{bar}$). The choice between them depends on power, response time, force density, environmental considerations, and cost.

Hydraulic power has high force density ($35\,\text{MPa} \times A$ gives kilonewtons from a small cylinder), fast response (essentially incompressible fluid, response limited by valves and lines), and is suited to heavy machinery (construction equipment, presses, aircraft control surfaces). Pneumatic power has moderate force density ($0.6$ to $1\,\text{MPa}$, an order of magnitude lower than hydraulics), low cost (compressed air is cheap to generate locally), and is suited to factory automation (clamping, sequencing, pick-and-place).

The basic equation of fluid power is the same as that of any mechanical power transmission: power equals force times velocity. In a hydraulic cylinder, force is pressure times area, velocity is flow rate divided by area, and power is pressure times flow rate. A $100\,\text{mm}$ cylinder at $200\,\text{bar}$ with flow rate $50\,\text{L/min}$ delivers $7.85\,\text{kN}$ at $0.106\,\text{m/s}$, or $830\,\text{W}$.

The efficiency of a hydraulic system has three components: the pump's efficiency (mechanical to fluid power, typically $80$ to $90\%$), the conduit losses (line friction and bend losses, typically $5\%$), and the valve and actuator losses (typically $10$ to $30\%$, dominated by control-valve pressure drops). The system efficiency is typically $50$ to $70\%$.

Pneumatic systems lose substantial energy in compressing the air ($\sim 50\%$ of the input power is dissipated as heat at the compressor) and in throttling (control valves drop pressure rather than convert pressure to motion, wasting the dropped energy). Pneumatic systems have efficiencies of $\sim 10\%$ overall; they are widely used despite this because the air is free, the components are inexpensive, and the infrastructure (compressed-air plant) is shared across many actuators.

The choice between hydraulic, pneumatic, and electric actuation follows the application. For sustained high force at modest speed (presses, lifts, mining equipment): hydraulics. For low to moderate force at fast cycles (factory automation, packaging): pneumatics. For precise positioning at moderate force (CNC, robotics, medical devices): electric. Many machines combine all three: a CNC lathe has hydraulic clamping, pneumatic chip removal, and electric drive.

\section{Hydraulic circuits, components, valves}\pathtag{standard}

A hydraulic circuit has a power source (a pump driven by a prime mover), a fluid distribution network (pipes and hoses), control valves, actuators (cylinders or rotary motors), filters, accumulators, reservoirs, and instrumentation.

Pumps deliver pressurised fluid. Fixed-displacement pumps (gear, vane, piston) deliver a constant flow proportional to speed; pressure is determined by the load. Variable-displacement pumps adjust their displacement under control to deliver only the flow required, saving energy. Variable-displacement pumps dominate modern hydraulic systems for energy efficiency.

Directional control valves direct fluid to actuators. Two-position, three-position, four-way valves with various centre positions (open centre, closed centre, tandem centre, float centre) implement the standard control patterns. Manually-operated, pilot-operated, and solenoid-operated valves cover the actuation methods.

Proportional and servo valves give continuous flow control rather than on-off. Proportional valves are less expensive and have moderate dynamic response (10 to $50\,\text{Hz}$ bandwidth); servo valves have higher precision and bandwidth ($100$ to $200\,\text{Hz}$) at greater cost. Servo valves dominate high-performance motion control (aerospace, machine tools, test rigs).

Pressure-relief valves protect against over-pressure (essential in any hydraulic circuit); pressure-reducing valves provide a regulated lower pressure for a sub-circuit; check valves enforce one-way flow; counterbalance valves prevent runaway under load on a falling actuator; flow control valves regulate flow rate.

Cylinders convert pressure into linear motion. Single-acting cylinders use pressure to extend and a spring or gravity to retract; double-acting cylinders use pressure for both directions. The cylinder's force is $F = P \cdot A$ (acting area), with the rod side area smaller than the cap side area, giving asymmetric force capability. Differential-area effects matter for symmetric motion.

Hydraulic rotary motors deliver torque proportional to pressure and displacement; speed is proportional to flow. Hydraulic motors are used where electric motors are unsuitable: heavy industrial equipment, drilling rigs, ship steering. Their power-to-mass ratio is high but their efficiency is lower than electric drives.

Filters, accumulators, and reservoirs are the supporting infrastructure. Filtration is essential: most hydraulic failures trace to contamination, with metallic and silicate particles destroying valve spools and pump elements. Accumulators store energy and absorb pressure spikes; reservoirs provide thermal dissipation and air separation. The reservoir is sized to allow the fluid to cool between passes; aggressive systems use heat exchangers.

\section{Pneumatic systems and applications}\pathtag{standard}

A pneumatic system has a compressor, an air drying and filtration unit, a distribution network, valves and actuators. The pressure is typically $6$ to $10\,\text{bar}$; the air is treated to remove water (which would condense in lines), oil (introduced by the compressor lubrication), and particulates.

Pneumatic cylinders and rotary actuators are similar to hydraulic but at lower pressure and with greater compressibility. The compressibility of air introduces dynamics: a loaded pneumatic cylinder oscillates when the valve opens, with the air volume acting as a spring. Critical control applications use pressure-feedback closed-loop control to address this.

Pneumatic logic was the original sequencing technology before electronic control became cheap; the discipline persists in some explosive environments (mining, oil-and-gas) where electrical sparking is prohibited. Pneumatic position controllers using bourdon tubes and nozzle-flapper amplifiers were the standard process-control technology of the 1950s and 1960s; pneumatic indicators and transmitters survive in hazardous-area applications.

Vacuum systems use sub-atmospheric pressure. Vacuum grippers (suction cups), vacuum chucks (holding parts on machine tables), and vacuum-assisted assembly are common pneumatic applications. The vacuum is generated by venturi (compressed air through a nozzle, simple) or by vacuum pump (efficient for larger loads).

Air bearings use a thin film of air to support a load with very low friction; the loads are modest (kg to tons) and the precision is high. Air bearings dominate semiconductor manufacturing (wafer stages with sub-nanometre accuracy), large optical instruments, and some precision measurement equipment.

\section{Combined systems and selection}\pathtag{mastery}

Real machines combine hydraulic, pneumatic, electric, and software subsystems. The selection of which subsystem performs which function follows the requirements: force, speed, precision, response, environment, cost, integration complexity, maintenance.

A construction-equipment hydraulic system: the engine drives a hydraulic pump; the pump feeds a manifold of valves that direct flow to cylinders for the boom, bucket, and travel; an electronic control unit reads operator commands (joystick, pedals) and modulates the valves; sensors give pressure, temperature, position. The pneumatic system handles brake actuation and air conditioning. The electric system handles lighting, instrumentation, and control. The architecture choice (load-sensing pump, variable-displacement pump, electro-hydraulic control) is a top-level engineering decision affecting fuel economy, response, and maintenance.

An aircraft flight-control system: hydraulic actuators move control surfaces under commands from the flight-control computer. Triple-redundant hydraulic systems (each with independent reservoir, pump, and lines) ensure that loss of one system does not lose control. The fly-by-wire architecture replaces mechanical control linkages with electrical signals to hydraulic servo valves; the safety argument requires diverse, redundant computers and actuators. The 787 transition to "more electric" (electric pumps and electric actuators for some functions) is incremental; full electric actuation of primary flight controls is not yet standard.

A machine-tool spindle: an electric motor (induction, PMSM, or BLDC) drives the spindle; the bearing system (often hybrid ceramic for high speed) handles the radial and axial loads; the cooling system (oil mist, liquid coolant) manages thermal effects; the lubrication system maintains the bearings; the control system tracks position and speed. The mechanical, electrical, fluid, and digital subsystems are all essential.

\section{Mechatronic integration: mechanical + electrical + digital}\pathtag{core}

Mechatronic integration is the discipline of combining mechanical, electrical, and digital subsystems into a working machine. The integration is harder than the sum of the subsystems because the interfaces between them introduce failure modes that no individual subsystem owns.

The interfaces between subsystems include physical interfaces (mechanical mounting, fluid connections, electrical connectors), electrical interfaces (signal levels, grounding, isolation, EMI), informational interfaces (data formats, protocols, timing), and operational interfaces (commissioning sequences, modes, fault responses). Each interface is a place where the design assumptions of one subsystem must align with the design assumptions of another; mismatches in assumption are the root cause of integration failures.

The integration architecture decision is made early and shapes the rest of the engineering. Centralised architectures put intelligence in a single controller with all sensors and actuators wired to it; the system is simple to reason about but the wiring is heavy and the controller is a single point of failure. Distributed architectures put intelligence at each node (smart sensors, smart actuators, smart subsystem controllers) communicating over a bus (CAN, EtherCAT, PROFINET); the wiring is lighter and the system is modular, but the verification surface is larger.

Modern industrial machines use distributed architectures: the PLC (programmable logic controller) or motion controller communicates with motor drives, I/O modules, smart sensors, vision systems, and HMI (human-machine interface) over fieldbuses. The standard bus depends on the industry (EtherCAT in machine tools, PROFINET in process industry, CAN in automotive, MIL-1553 in aerospace).

The integration discipline includes: defining the interfaces clearly and early, qualifying each subsystem against its interface specification, testing the integrated system thoroughly, documenting the system as built, and providing tools for diagnosis and maintenance. The integration engineer's most useful artefact is the system diagram showing every subsystem, every interface, every signal, every fluid line; the integration engineer's most useful skill is the discipline of treating each interface as a specification problem.

\begin{archetype}[interface: integration as interface management]
A mechatronic system fails at interfaces more often than at components. The motor works; the encoder works; the controller works; but the encoder cable picks up noise from the motor drive, the controller reads garbage, and the motion is wrong. The valve works; the controller works; but the controller's command range and the valve's accepted range do not match, and the system saturates early. The pump works; the heat exchanger works; but the line between them has a smaller diameter than specified, the flow restricts, and the system overheats. The integration engineer's job is interface management: enumerate every interface, specify every interface, qualify every interface, test every interface.
\end{archetype}

\section{System-level design for the full machine}\pathtag{standard}

System-level design considers the machine as a whole, decomposing requirements into subsystem specifications and verifying that the subsystem specifications collectively meet the system requirements. The standard tools are requirements traceability (every requirement maps to a verification method), interface control documents (every interface has a specification), system FMEA (every failure mode maps to a response), and system-level testing (every requirement is verified on the integrated system).

The system architect's first decision is the partition: what each subsystem does, what each interface carries, what each component is responsible for. The partition determines what is easy and what is hard for the rest of the engineering. A clean partition with well-defined interfaces makes development modular and verification tractable; a tangled partition with overlapping responsibilities makes everything harder.

The next decision is the standards basis: which industry standards apply (machine safety EN 62061 / ISO 13849, electrical safety IEC 61010 / UL 508, EMC IEC 61000 series, regulated environments medical IEC 60601 / aerospace DO-178/254 / automotive ISO 26262). The standards prescribe development processes, documentation, testing, and safety architectures. Choosing the standards basis early is cheaper than retrofitting compliance.

The third decision is the verification strategy: simulation, prototyping, scaling, qualification. Modern complex machines use model-based engineering: a system simulation that includes the major dynamics is built in parallel with the hardware, used to verify the design and to support control development. Hardware-in-the-loop (HIL) testing connects real hardware (sensors, actuators, controllers) to a simulated environment, enabling tests that would be expensive or dangerous on the real machine.

The fourth decision is the manufacturing and supply chain: what is in-house, what is outsourced, what is custom, what is off-the-shelf. The decision drives lead times, cost, intellectual-property considerations, and supplier dependence.

\section{Documentation, commissioning, handover}\pathtag{standard}

A machine is not delivered when the hardware is built and the firmware is loaded; it is delivered when the customer or operator can run, maintain, and depend on it. The engineering work between "powers on" and "customer accepts" is the commissioning phase, and it is supported by documentation.

The documentation set typically includes: system description (what the machine does), specifications (numerical performance targets), drawings (mechanical, electrical, hydraulic, pneumatic), bill of materials, theory of operation (how the machine works), operating procedures (start-up, normal operation, shutdown), maintenance procedures (preventive and corrective), spare parts list, software documentation (functional specification, design documentation, source code with comments, build instructions, version control), and certification documentation (test reports, compliance evidence, traceability).

Commissioning is the systematic process of bringing the machine into operation. The standard sequence is: pre-power-on checks (wiring continuity, fluid connections, no foreign objects), power-on with reduced energy (low pressure or low voltage), subsystem-by-subsystem verification, integrated operation at reduced parameters, full-parameter operation, performance verification against specification, training of operators, sign-off.

Handover to the customer transfers responsibility for the machine to the operating organisation. The handover should include the documentation, training, initial spares, and a defined warranty and support agreement. A clean handover reduces follow-up support burden.

\begin{principle}[The commissioning-and-documentation principle]
A machine that works once does not constitute engineering. Engineering is the discipline that makes the machine work reliably, safely, and reproducibly, in the hands of operators who did not design it, over the machine's full lifetime. Documentation, commissioning procedures, and handover are not afterthoughts; they are the artefacts that allow others to depend on the engineered machine. Skip them, and the engineer's work is fragile.
\end{principle}

\section{Failure: leak failures, contamination, integration interface failures}\pathtag{core}

Integrated machines fail in characteristic ways.

\textbf{Leak failures:} a hydraulic or pneumatic seal degrades, a hose ruptures, a fitting loosens. Symptoms: fluid loss, pressure loss, contamination of the environment, reduced performance. Hydraulic leaks are especially serious: high-pressure fluid streams can injure (injection injuries) and ignite (oil mist explosions). Mitigation: quality fittings, proper torque, contamination control during assembly, regular inspection, hose-burst protection (valves that close on pressure drop).

\textbf{Contamination:} particles in the fluid (metal wear, silicates, dust ingress, water condensation) damage valves and pumps. The cleanliness class of the fluid (ISO 4406 / NAS 1638) sets the contamination tolerance; aerospace and high-performance systems require lower contamination than industrial systems. Mitigation: filtration (with sized retention), reservoir design, contamination ingress control, periodic fluid testing and replacement.

\textbf{Integration interface failures:} sensors that produce values the controller does not expect; actuators that respond outside their commanded range; communication protocols where the parties disagree on the message format; timing assumptions that fail in production conditions; calibrations that drift away from the as-commissioned values. Mitigation: interface specifications, interface testing, integration testing, condition monitoring, regular calibration.

\textbf{Mechanical wear and fatigue:} the mechanical subsystem degrades with use (covered in Chapter 4); the integrated machine inherits these failure modes plus integration-specific patterns (a worn linkage produces position error that the controller corrects for, masking the wear until it is severe).

\textbf{Software bugs in integrated firmware:} the integration software handles multiple subsystems and their interactions; bugs in this layer present as system-level failures with no obvious subsystem root cause. The Therac-25 race condition is the canonical example: the electron-beam current selection and the beam-shaping plate position were both controlled by software; a fast operator sequence left them in an inconsistent state. Mitigation: integration-level testing, design review, formal methods for safety-critical interactions.

\textbf{Operator and procedural failures:} the machine is operated outside its intended envelope, or the operator's mental model of the machine does not match its behaviour. Mitigation: training, clear documentation, conservative operating limits enforced in software, error-tolerant interface design.

\textbf{Field-environment failures:} the machine works in the lab and fails in service. Temperature extremes, vibration, humidity, dust, contaminants, electromagnetic interference, power-quality variation can all degrade integrated performance. Mitigation: environment qualification testing (IEC 60068 climatic series, IEC 60721 environmental classes), conservative design margins, field validation under realistic conditions.

The 1986 Challenger O-ring failure was an integration failure: the solid-rocket booster joint's seal performance at cold temperatures was outside the qualified range; the launch decision did not adequately weight the warning. The 2003 Columbia foam-strike failure was an integration failure: the external tank's foam insulation could shed under launch conditions; the impact damaged the wing leading-edge thermal protection; the in-flight inspection capability was missing. Both accidents are integration-engineering case studies as much as they are aerospace case studies.

\section{Project}\pathtag{core}

\begin{project}[Mechatronic system: mechanical, electrical, digital, fluid; integrate; document]
\textbf{Objective.} Design, build, integrate, document, and demonstrate a small mechatronic system that includes a mechanical structure, electrical actuation, digital control, and a fluid (hydraulic or pneumatic) subsystem. Document the system to the standard that a peer could replicate it.

\textbf{Method.} Choose a project that integrates all four subsystems. Options: a pneumatic-actuated stamping press with a microcontroller-based sequencer, encoder feedback, and a digital safety circuit; a small hydraulic positioner (using a syringe pump or laboratory-scale hydraulic kit) under digital position control; a robotic gripper with pneumatic actuation, force sensing, and digital control. Design each subsystem; specify each interface; build, test, integrate, and commission. Document the system: schematic (electrical), hydraulic or pneumatic circuit diagram, mechanical drawings, firmware listing, theory of operation, operating procedure, maintenance procedure, and a verification report showing each requirement is met.

\textbf{Deliverables.} The hardware (assembled and operational), the documentation set (drawings, source code, theory of operation, operating procedure), a demonstration video, and a written reflection on the integration challenges encountered: what interface design decisions worked, what did not, what would be done differently.

\textbf{Hazard.} Hydraulic systems at pressure can inject fluid through skin (injection injury, medical emergency); pneumatic systems can produce high-velocity flying parts on rupture; electrical hazards from supplies; mechanical hazards from moving parts. Use lab-scale fluid systems (low pressure); wear safety glasses; engineer end-stops and emergency-stops into the design.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A hydraulic cylinder has bore $80\,\text{mm}$ and operates at $200\,\text{bar}$. Compute the force on the cap side. The rod is $30\,\text{mm}$; compute the force on the rod side.
\end{exercise}

\begin{exercise}
A hydraulic pump delivers $40\,\text{L/min}$ at $180\,\text{bar}$. Compute the hydraulic power delivered (in $\text{kW}$) and the input power required if the pump efficiency is $88\%$.
\end{exercise}

\begin{exercise}
A pneumatic cylinder of $50\,\text{mm}$ bore operates at $6\,\text{bar}$. Compute the extension force.
\end{exercise}

\begin{exercise}
A hydraulic line has $30\,\text{mm}$ internal diameter, $5\,\text{m}$ length, oil viscosity $40\,\text{cSt}$, flow $80\,\text{L/min}$. Estimate the pressure drop (use Darcy-Weisbach or appropriate model).
\end{exercise}

\begin{exercise}
A reservoir holds $200\,\text{L}$ of hydraulic oil; the system flow is $100\,\text{L/min}$ at $250\,\text{bar}$. Compute the residence time in the reservoir and assess whether it is adequate for thermal dissipation.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the relationship between flow rate, valve orifice area, and pressure drop for a hydraulic control valve. State the conditions for cavitation.
\end{exercise}

\begin{exercise}
Derive the natural frequency of a pneumatic cylinder-mass system with air volume on each side.
\end{exercise}

\begin{exercise}
Derive the efficiency relationship between hydraulic and electrical actuation for a given linear motion task with sustained holding force.
\end{exercise}

\begin{exercise}
Derive the volumetric efficiency of a hydraulic pump as a function of pressure, including internal leakage.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the energy cost of operating a $7\,\text{bar}$ pneumatic system at $0.5\,\text{Nm}^3/\text{min}$ continuous demand, with electricity at $\$0.15/\text{kWh}$ and a compressor efficiency of $30\%$.
\end{exercise}

\begin{exercise}
Estimate the size of a hydraulic accumulator required to deliver $10\,\text{L}$ at $200\,\text{bar}$ for a peak-power event, with a precharge of $100\,\text{bar}$.
\end{exercise}

\begin{exercise}
Estimate the cleanliness requirement (ISO 4406 class) for a hydraulic system with servo valves; compare to a system with proportional valves only.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate the position-control loop of a hydraulic actuator with servo valve, including fluid compressibility. Identify the control bandwidth.
\end{exercise}

\begin{exercise}
Simulate the integration timing of a mechatronic system: $1\,\text{kHz}$ control loop, $100\,\text{Hz}$ CAN messages, $30\,\text{Hz}$ vision update. Identify the worst-case end-to-end latency.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design the actuation architecture for a $1\,\text{ton}$ shop press for forming sheet metal. Compare hydraulic, electric servo, and pneumatic options.
\end{exercise}

\begin{exercise}
Design the integration architecture (bus selection, partitioning, fault response) for a CNC machine with $5$ axes, spindle, ATC, chip-management, and operator interface.
\end{exercise}

\begin{exercise}
Design the documentation set required for handover of an industrial machine to a contract manufacturer who must operate and maintain it without ongoing support.
\end{exercise}

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A team's hydraulic system shows declining response over months. State three engineering hypotheses (contamination, wear, leak), the diagnostic measurements for each, and the corresponding maintenance actions.
\end{exercise}

\begin{exercise}
A team is choosing between hydraulic and electric actuation for the boom of a small excavator. State the engineering trade-offs.
\end{exercise}

\begin{exercise}
A team's machine works in the factory and fails at a customer's site. State three engineering hypotheses related to environmental differences.
\end{exercise}

\begin{exercise}
A team is asked to add a software feature to an industrial controller that was certified to IEC 61508 SIL 2. State the engineering implications and the development-process requirements.
\end{exercise}

\begin{exercise}
A team's integrated mechatronic system has interface failures between the vision module and the motion controller. State the engineering response and the role of interface specifications.
\end{exercise}

\begin{exercise}
A team is commissioning a new machine; intermittent faults appear under specific operating conditions. State the engineering troubleshooting approach.
\end{exercise}

\begin{exercise}
A team is asked to retrofit modern digital control onto a 30-year-old hydraulic machine. State the engineering considerations: sensor selection, controller integration, safety, documentation, and the conditions under which a full replacement is preferred over a retrofit.
\end{exercise}

\begin{exercise}
A team's machine documentation has fallen out of step with the as-built hardware after several rounds of field modifications. State the engineering response and the discipline required to bring documentation back into agreement with reality.
\end{exercise}
