\chapter{Reactions and equilibria}
\label{vol05:ch02}

\epigraph{Every chemical action without exception can take place in two
opposite directions.}{Henri Le Chatelier, paraphrased in
\textcite[ch.~6]{text:atkins-physical-chemistry}, the standard
working reference for the modern statement of the principle.}

\chapmeta{Half-life: foundational. Archetypes: balance, stability, failure. Exercise target: 28. Project track: household analytical chemistry (acid-base titration). Hazard class: low (kitchen-scale dilute acid and base). Reader path default: standard, with sections explicitly tagged. Named cases: Bhopal methyl isocyanate release (1984).}

\noindent
A chemical reaction is the rearrangement of atoms among bonds.
Volume IV gave the thermodynamic bookkeeping that decides which
arrangements are energetically allowed; the present chapter turns the
bookkeeping into the working vocabulary of reactions, rates,
equilibria, and the failure modes that arise when an exothermic
reaction's heat generation outruns the system's ability to remove it.
The chapter establishes stoichiometry, the Arrhenius law, the
equilibrium constant and Le Chatelier's principle, the four classes of
common reactions (acid-base, redox, complexation, precipitation), the
free-energy form of the equilibrium criterion, and the catalysis that
short-circuits the rate without changing the equilibrium. The closing
section names the bond-level mechanism by which loss of cooling, loss
of containment, or loss of mixing control leads to the most consequential
failure mode in process engineering, the runaway reaction.

\section{Stoichiometry refresher}\pathtag{core}

A chemical equation is an accounting identity for atoms. The two sides
of the equation must balance in every chemical element and in net
electrical charge. The combustion of methane,
\[
\text{CH}_4 + 2\,\text{O}_2 \to \text{CO}_2 + 2\,\text{H}_2\text{O},
\]
balances one carbon, four hydrogens, and four oxygens on each side.
The coefficients ($1$, $2$, $1$, $2$ here) are the
\engterm{stoichiometric coefficients}; they fix the mole ratios in
which the reactants combine and the products appear. The equation says
nothing about rate, mechanism, or whether the reaction proceeds; it
says only that whenever it does proceed, the amounts of substance
involved follow the ratios shown.

\engterm{The mole}. The amount of substance $n$ containing $N_A = 6.022
\times 10^{23}$ specified entities. The mole is the chemist's working
unit; it carries the bridge between the atomic-scale picture of
Chapter~1 and the laboratory-bench amounts of solid, liquid, or gas.
The mass of one mole of a substance is its \engterm{molar mass} $M$,
numerically equal to the atomic or molecular mass expressed in atomic
mass units. The molar mass of methane is $16.04\,\text{g}/\text{mol}$;
of oxygen O$_2$, $32.00\,\text{g}/\text{mol}$; of carbon dioxide,
$44.01\,\text{g}/\text{mol}$; of water, $18.02\,\text{g}/\text{mol}$.
The stoichiometric coefficients combined with the molar masses give
the mass ratios: one gram of methane combines with $4\,\text{g}$ of
oxygen and produces $2.75\,\text{g}$ of carbon dioxide and $2.25\,
\text{g}$ of water.

\engterm{Limiting reactant}. When two or more reactants are mixed in
non-stoichiometric proportions, the reactant present in deficiency
relative to the stoichiometric ratio is the \engterm{limiting
reactant}; it is consumed completely if the reaction goes to
completion, and the excess of the other reactants remains. The
limiting reactant sets the maximum yield of product. An engineer
designing a batch reactor identifies the limiting reactant, computes
the theoretical yield, and compares to the measured yield to obtain
the conversion.

\engterm{Conversion and yield}. The \engterm{conversion} $X$ of a
reactant is the fraction of its initial amount that has been
consumed. The \engterm{selectivity} of a product is the fraction of
the consumed reactant that has been converted to that specific
product, when multiple products are possible. The \engterm{yield} is
the product of the conversion and the selectivity. A reaction with
$X = 0.9$ conversion and $0.95$ selectivity to the desired product
yields $0.855$ of the theoretical maximum. The three quantities are
the working metrics of process performance in chemical engineering.

\engterm{Extent of reaction}. A reaction's progress is conveniently
described by a single scalar variable, the extent $\xi$, defined so
that the change in moles of any species $i$ during the reaction is
$\Delta n_i = \nu_i \xi$, where $\nu_i$ is the stoichiometric
coefficient (negative for reactants, positive for products). The
extent has units of moles. For the methane combustion above, $\xi =
1\,\text{mol}$ corresponds to consuming one mole of CH$_4$ and two
moles of O$_2$ to produce one mole of CO$_2$ and two moles of
H$_2$O. The extent is the engineer's bookkeeping variable for a
single reaction; for a network of reactions, one extent variable per
independent reaction.

\engterm{Mass balance}. For a closed reacting system, the total mass
is conserved (an exact consequence of conservation of atoms and the
mass-energy relation in non-relativistic chemistry). For an open
reacting system (a control volume with mass flow), the species mass
balance for species $i$ in the steady state is
\[
\dot{m}_{i,\text{in}} - \dot{m}_{i,\text{out}} + \dot{m}_{i,\text{gen}} = 0,
\]
where the generation rate $\dot{m}_{i,\text{gen}}$ is positive for a
species produced by reaction and negative for a species consumed.
Process engineers write one such balance per species and one overall
mass balance and use the system to size reactors and identify
non-physical inputs.

\begin{archetype}[Balance]
The chemical-reaction balance is identical in form to the energy and
momentum balances of Volumes III and IV: a generation term, an inflow,
an outflow, and an accumulation. The accounting discipline is the
same. The species generation term is non-trivial: it depends on the
rate law and on the local composition. The balance is the engineer's
instrument for designing reactors, sizing storage, predicting yields,
and identifying unaccounted byproducts.
\end{archetype}

\section{Reaction rates and Arrhenius}\pathtag{core}

A reaction's \engterm{rate} is the rate of change of extent per unit
volume per unit time. For a homogeneous reaction in a well-mixed
system at constant volume, the rate is $r = (1/V)(d\xi/dt)$. For a
single reaction with rate law
\[
r = k \prod_i c_i^{\alpha_i},
\]
where $c_i$ is the concentration of species $i$ (typically moles per
litre) and $\alpha_i$ is the order with respect to that species, the
rate constant $k$ has units that depend on the overall order. A
first-order reaction has $k$ with units $\si{\per\second}$; a
second-order reaction, $\text{L}/(\text{mol}\cdot\si{\second})$.

The order with respect to each species is determined empirically. For
elementary reactions (those that proceed in a single bond-rearrangement
step) the order equals the stoichiometric coefficient of the reactant;
for non-elementary reactions, the order may be fractional or
non-integer, reflecting a more complicated mechanism. The hydrolysis
of ethyl acetate by aqueous base is first-order in both ester and
hydroxide and second-order overall, with rate constant $k \approx
0.11\,\text{L}/(\text{mol}\cdot\si{\second})$ at $25\,\degC$; the
decomposition of nitrogen pentoxide in the gas phase is first-order,
with $k \approx 3 \times 10^{-5}\,\si{\per\second}$ at $25\,\degC$
\cite[ch.~17]{text:atkins-physical-chemistry}.

\engterm{The Arrhenius equation}. The temperature dependence of nearly
every reaction rate constant is well described by the empirical form
\[
k(T) = A \exp\left(-\frac{E_a}{R T}\right),
\]
introduced by Arrhenius in 1889 \cite{hist:arrhenius1889}, where $A$
is the \engterm{pre-exponential factor} (with the same units as $k$),
$E_a$ is the \engterm{activation energy} (joules per mole), $R =
8.314\,\text{J}/(\text{mol}\cdot\si{\kelvin})$ the gas constant, and
$T$ the absolute temperature. The activation energy is the height of
the energy barrier the reactants must climb over the transition state
to become products; the pre-exponential factor is approximately the
frequency of collisions with the right geometry. For typical
solution-phase organic reactions, $E_a$ is in the range $20$ to $200\,
\text{kJ}/\text{mol}$; for combustion reactions of hydrocarbons,
$E_a$ is in the range $100$ to $400\,\text{kJ}/\text{mol}$.

A rule of thumb that follows from the Arrhenius form is that a
$10\,\si{\kelvin}$ increase in temperature roughly doubles the rate
of a typical organic reaction with $E_a \approx 50\,\text{kJ}/\text{mol}$
near room temperature. The doubling factor is precise enough for a
back-of-envelope check on the plausibility of a kinetic data sheet.

\engterm{Reaction order from data}. For a reaction $A \to B$ with
rate $r = k c_A^n$ in a batch reactor of constant volume, the
integrated rate law is
\[
n = 1: \quad c_A(t) = c_A(0) \exp(-k t),
\]
\[
n = 2: \quad \frac{1}{c_A(t)} - \frac{1}{c_A(0)} = k t,
\]
\[
n \neq 1: \quad c_A(t)^{1-n} - c_A(0)^{1-n} = (n-1) k t.
\]
A first-order reaction shows linear $\ln c_A$ vs $t$; a second-order
reaction, linear $1/c_A$ vs $t$. The reader who plots both
transformations on the same data set identifies the order from which
transformation gives the straighter line. The \engterm{half-life} of a
first-order reaction is $t_{1/2} = \ln 2/k$, independent of initial
concentration; for second-order, $t_{1/2} = 1/(k c_A(0))$, increasing
as $c_A(0)$ decreases.

\section{Chemical equilibrium and Le Chatelier}\pathtag{core}

Every chemical reaction is, in principle, reversible. The reverse
reaction has its own rate; at equilibrium, the forward and reverse
rates are equal and the system's composition is steady. The
\engterm{equilibrium constant} $K$ for the reaction
\[
aA + bB \rightleftharpoons cC + dD
\]
is
\[
K = \frac{a_C^c \, a_D^d}{a_A^a \, a_B^b},
\]
where $a_i$ is the \engterm{activity} of species $i$ at equilibrium.
For ideal gases, the activity is the partial pressure divided by a
standard pressure ($a_i = p_i/p^\circ$, with $p^\circ = 1\,\text{bar}$
the standard); for ideal solutions, the concentration divided by a
standard concentration ($a_i = c_i/c^\circ$, with $c^\circ = 1\,
\text{mol}/\text{L}$). The equilibrium constant is dimensionless when
defined this way; many textbooks write it with implied units, which
the engineer must reconstruct from the standard states.

The equilibrium constant depends on temperature through the
van't Hoff equation,
\[
\frac{d \ln K}{d T} = \frac{\Delta H^\circ}{R T^2},
\]
which integrates to
\[
\ln\frac{K(T_2)}{K(T_1)} = -\frac{\Delta H^\circ}{R}\left(\frac{1}{T_2}
- \frac{1}{T_1}\right).
\]
For an exothermic reaction ($\Delta H^\circ < 0$), $K$ decreases with
increasing temperature: the equilibrium shifts toward reactants. For
endothermic reactions, $K$ increases with $T$. The qualitative
content is \engterm{Le Chatelier's principle}: a system at equilibrium
responds to a perturbation by shifting its composition in the direction
that opposes the perturbation. Higher temperature pushes an
exothermic equilibrium back to reactants (the system absorbs heat by
running the reaction in the endothermic direction); higher pressure
pushes a reaction toward the side with fewer moles of gas; addition
of a product shifts the reaction back toward reactants.

\engterm{Reaction quotient}. The same expression as $K$ but evaluated
at the current (non-equilibrium) composition: $Q = \prod a_i^{\nu_i}$.
The direction of net reaction is set by the sign of $\ln(Q/K)$: if
$Q < K$, the reaction proceeds forward (toward products); if $Q > K$,
the reverse; if $Q = K$, the system is at equilibrium. The reaction
quotient is the engineer's diagnostic for whether a given mixture
will react and in which direction.

\engterm{Heterogeneous equilibria}. When the reactants and products
are in different phases, the activity of a pure condensed phase
(pure solid, pure liquid) is taken as unity, and only the gas-phase
or solution-phase species appear in the equilibrium expression. The
decomposition of calcium carbonate to calcium oxide and carbon
dioxide,
\[
\text{CaCO}_3(s) \rightleftharpoons \text{CaO}(s) + \text{CO}_2(g),
\]
has equilibrium expression $K = p_{\text{CO}_2}/p^\circ$. The
equilibrium $p_{\text{CO}_2}$ depends only on temperature; at
$900\,\degC$ it is approximately $1\,\text{bar}$, which is why
cement kilns operate above that temperature.

\engterm{Solubility products}. For a sparingly soluble ionic solid
$M_pX_q$ in water, the equilibrium with its dissolved ions,
\[
M_pX_q(s) \rightleftharpoons p\,M^{q+}(aq) + q\,X^{p-}(aq),
\]
has equilibrium constant $K_{sp} = [M^{q+}]^p [X^{p-}]^q$, where the
brackets denote dimensionless activities (concentrations divided by
$c^\circ$). The solubility product is the engineer's working number
for predicting whether a solution will precipitate; it is the
quantitative form of the qualitative ``insoluble'' label on a
chemistry chart.

\begin{principle}[Equilibrium is a state, not a stopping point]
A reaction at equilibrium has zero net rate of composition change but
non-zero forward and reverse rates. The composition is fixed by the
equilibrium constant; the rate of approach to equilibrium is set by
the forward and reverse rate constants. The two pieces (where the
system goes, and how fast) are independent. The engineer who confuses
the equilibrium position with the reaction rate has confused the
destination with the speed.
\end{principle}

\section{Acid-base, redox, complexation}\pathtag{standard}

Four broad classes of reaction dominate engineering chemistry:
acid-base, redox, complexation, and precipitation. The classes are
distinguished by what transfers from reactant to product (a proton, an
electron, a ligand, an ion pair, respectively); each has its own
working equilibrium constant and its own engineering applications.

\engterm{Acid-base reactions}. A \engterm{Bronsted acid} is a proton
donor; a \engterm{Bronsted base} is a proton acceptor. The
self-ionisation of water,
\[
2\,\text{H}_2\text{O} \rightleftharpoons \text{H}_3\text{O}^+ +
\text{OH}^-,
\]
has equilibrium constant $K_w = [\text{H}_3\text{O}^+][\text{OH}^-] =
1.0 \times 10^{-14}$ at $25\,\degC$. The dimensionless quantity
\engterm{pH} is defined as $\text{pH} = -\log_{10}[\text{H}_3\text{O}^+]$;
at $25\,\degC$, pure water has $[\text{H}_3\text{O}^+] = 10^{-7}\,
\text{mol}/\text{L}$ and pH $= 7$. Acidic solutions have pH $< 7$;
basic solutions, pH $> 7$.

An acid HA in water dissociates with equilibrium constant $K_a$,
\[
\text{HA} \rightleftharpoons \text{H}^+ + \text{A}^-, \quad K_a =
\frac{[\text{H}^+][\text{A}^-]}{[\text{HA}]}.
\]
Strong acids (HCl, HNO$_3$, H$_2$SO$_4$) have $K_a$ very large, so
they are essentially fully dissociated; their effective concentration
of $\text{H}^+$ equals the formal concentration of the acid. Weak
acids (acetic acid, $K_a = 1.8 \times 10^{-5}$; benzoic acid, $K_a =
6.5 \times 10^{-5}$; carbonic acid, $K_a = 4.3 \times 10^{-7}$) are
only partially dissociated. The $\text{p}K_a = -\log_{10} K_a$ of a
weak acid is the pH at which the acid is half-dissociated; the
\engterm{Henderson-Hasselbalch equation} $\text{pH} = \text{p}K_a +
\log_{10}([\text{A}^-]/[\text{HA}])$ describes a buffer solution.

\engterm{Buffers}. A solution of a weak acid and its conjugate base in
comparable concentrations resists changes in pH when small amounts of
strong acid or base are added; the conjugate pair absorbs the proton
or hydroxide. Buffers are the working pH-control mechanism in
biological systems (blood at pH $7.4$, buffered by the
bicarbonate-carbonic-acid pair) and in laboratory and process
chemistry (acetate buffers at pH $\sim 5$, phosphate buffers at
pH $\sim 7$, ammonia buffers at pH $\sim 9$). The reader who chooses
a buffer for an engineering application picks one whose p$K_a$ is
within one unit of the target pH.

\engterm{Redox reactions}. \engterm{Oxidation} is loss of electrons;
\engterm{reduction} is gain of electrons. An \engterm{oxidising agent}
gains electrons; a \engterm{reducing agent} loses them. In the rusting
of iron,
\[
4\,\text{Fe} + 3\,\text{O}_2 \to 2\,\text{Fe}_2\text{O}_3,
\]
iron loses electrons (oxidation) and oxygen gains them (reduction).
The reaction can be split into half-reactions:
\[
\text{Fe} \to \text{Fe}^{3+} + 3 e^-, \quad \text{O}_2 + 4 e^- + 2\,
\text{H}_2\text{O} \to 4\,\text{OH}^-.
\]
The two half-reactions can be physically separated in an
electrochemical cell, the topic of Chapter~3 of this volume. The
oxidation state of an atom in a compound is a bookkeeping device: the
charge the atom would carry if every bond were fully ionic, with the
electrons assigned to the more electronegative partner. In iron
oxide Fe$_2$O$_3$, iron has oxidation state $+3$; in iron sulfide
FeS$_2$ (pyrite), $+2$.

\engterm{Complexation}. A complex is a Lewis-acid-Lewis-base
combination: a central metal cation (Lewis acid) binds to one or more
ligands (Lewis bases, electron-pair donors). The complexes of
transition metals are responsible for the colours of solutions
(chromium-aquo complexes are green or violet, copper-aquo blue,
iron-aquo pale yellow), for the chelating behaviour of EDTA
(ethylenediaminetetraacetic acid, used to sequester polyvalent metal
ions in water-softening and in food preservation), for the activity
of metalloenzymes in biology, and for the function of catalytic
metal centres in industrial chemistry. The equilibrium constant for
the formation of a complex from its components is the
\engterm{formation constant} $K_f$; values range from $10^2$ to
$10^{20}$ across the complex literature.

\engterm{Precipitation}. The reverse of dissolution. A solution
supersaturated in a poorly soluble ionic compound precipitates the
solid until the remaining solution satisfies the $K_{sp}$ relation.
Industrial precipitation is the working operation in water treatment
(removal of hardness ions by addition of sodium carbonate to
precipitate calcium and magnesium carbonates), in mineral processing
(precipitation of metals from leach solutions), and in pharmaceutical
crystallisation. The engineering issue with precipitation is usually
crystal size and habit, not the equilibrium yield; the supersaturation
and the nucleation kinetics set the particle distribution that
emerges.

\section{Free energy revisited}\pathtag{standard}

Volume IV's Gibbs free energy $G$ is the working equilibrium criterion
of chemistry at constant temperature and pressure. The change in
Gibbs free energy for a reaction at the standard state,
\[
\Delta G^\circ = \Delta H^\circ - T \Delta S^\circ,
\]
combines the enthalpy change (heat released or absorbed at constant
pressure) with the entropy change (disorder change) into a single
spontaneity criterion. A reaction with $\Delta G^\circ < 0$ at the
standard state has equilibrium constant $K > 1$ and proceeds toward
products at standard composition; a reaction with $\Delta G^\circ > 0$
has $K < 1$ and proceeds toward reactants. The quantitative link is
\[
\Delta G^\circ = -RT \ln K,
\]
which converts standard free-energy changes (tabulated for thousands
of compounds) into equilibrium constants directly.

\engterm{The reaction free energy at non-standard composition}. The
actual driving force for reaction at composition with reaction
quotient $Q$ is
\[
\Delta G = \Delta G^\circ + RT \ln Q.
\]
At equilibrium $\Delta G = 0$ and $Q = K$, recovering the relation
above. The non-equilibrium expression is the working form used in
electrochemistry (Chapter~3 of this volume), where the cell voltage is
$\Delta G/(nF)$ with $n$ the number of electrons transferred per
mole and $F$ Faraday's constant.

\engterm{The temperature dependence}. From van't Hoff, the equilibrium
constant rises with temperature for endothermic reactions and falls
for exothermic. The engineer who wishes to push an equilibrium to
the product side picks the temperature accordingly: ammonia synthesis
($\text{N}_2 + 3\text{H}_2 \rightleftharpoons 2\text{NH}_3$, $\Delta
H^\circ = -92\,\text{kJ}/\text{mol}$) is exothermic, so equilibrium
yield is highest at low temperature, but the rate is too slow there;
the Haber-Bosch process compromises by running at $400$ to $500\,
\degC$ and $150$ to $300\,\text{bar}$, with an iron catalyst. The
equilibrium yield at the high temperature is only $15$ to $20\,\%$,
recovered by recycling unreacted gas. Chapter~5 of Volume IV gives
the cycle analysis; the present chapter gives the equilibrium
analysis.

\engterm{The Gibbs energy of mixing}. When two components mix to form
an ideal solution at constant $T$ and $p$, the change in Gibbs energy
is
\[
\Delta G_{\text{mix}} = nRT(x_A \ln x_A + x_B \ln x_B) < 0,
\]
with mole fractions $x_A$ and $x_B$. The mixing is spontaneous for any
non-zero composition because the entropy of mixing is positive and
the ideal enthalpy of mixing is zero. Real solutions show non-ideal
enthalpy of mixing, sometimes large enough to drive phase separation
(Chapter~4 of this volume develops the phase-diagram consequences).
The mixing Gibbs energy is also the working quantity in osmotic
pressure, in the cryoscopic and ebullioscopic constants, and in
colligative properties more broadly (Chapter~3).

\section{Catalysis}\pathtag{standard}

A \engterm{catalyst} is a substance that increases the rate of a
reaction without being consumed by it and without changing the
reaction's equilibrium position. The catalyst provides an alternative
reaction pathway with a lower activation energy; the same equilibrium
constant is approached more quickly. The number $K$ does not change;
the rate constants $k_+$ and $k_-$ both increase by the same factor,
preserving $K = k_+/k_-$.

The engineering uses of catalysis are pervasive. The Haber-Bosch
process for ammonia synthesis uses iron-based catalysts (with
potassium and aluminium oxide promoters) that lower the activation
energy of the dissociative chemisorption of N$_2$ on the catalyst
surface, which is the rate-limiting step. The contact process for
sulphuric acid uses vanadium pentoxide to oxidise sulphur dioxide
to sulphur trioxide. The catalytic cracker in a petroleum refinery
uses zeolite catalysts to break large hydrocarbon molecules into
smaller ones at temperatures and pressures lower than would otherwise
be needed. Automobile three-way catalytic converters use platinum,
palladium, and rhodium on a ceramic monolith to oxidise carbon
monoxide and unburned hydrocarbons while reducing nitrogen oxides;
the reactions run rapidly enough at exhaust-gas temperatures to clean
the gas stream during its $0.1$ to $0.5\,\si{\second}$ residence time
in the converter.

\engterm{Homogeneous vs heterogeneous catalysis}. A \engterm{homogeneous}
catalyst is in the same phase as the reactants (typical of acid-base
catalysis in solution and of organometallic catalysis); a
\engterm{heterogeneous} catalyst is in a different phase (typical of
gas-phase reactions catalysed by solid surfaces, of liquid-phase
reactions catalysed by suspended solids). Heterogeneous catalysis is
the dominant mode in industrial chemistry because the catalyst can be
recovered from the product stream by filtration or settling.
Homogeneous catalysis often gives higher selectivity at the cost of
harder separation.

\engterm{Enzymes}. The biological catalysts of every living cell. An
enzyme is a protein (occasionally an RNA molecule) that binds a
specific substrate in a specific geometric orientation and lowers the
activation energy of a specific reaction. The catalytic efficiencies
are extreme: carbonic anhydrase, the enzyme that interconverts CO$_2$
and bicarbonate in blood, has turnover number $10^6\,\si{\per\second}$
near the diffusion limit. The temperature dependence of enzyme
catalysis usually follows Arrhenius below the protein's thermal
denaturation temperature; above that temperature the protein unfolds
and catalysis stops. Industrial enzyme catalysis (proteases in
detergents, amylases in baking, glucose isomerase in high-fructose
corn syrup) operates near the thermal optimum of the chosen enzyme.

\engterm{Catalyst poisoning and deactivation}. Real catalysts lose
activity in service. The loss mechanisms are well catalogued:
chemical poisoning (a foreign species that binds the active site
more strongly than the reactant, blocking it), thermal sintering
(the catalyst surface coarsens at high temperature, reducing the
active area), coking (hydrocarbon residues build up on the surface),
and mechanical attrition (the catalyst particles break up under the
flow). Engineering catalysis includes the design of the regeneration
or replacement cycle that maintains activity over the plant's
operating life. Automotive catalytic converters lose activity to
lead in fuel (the reason for unleaded fuel in the post-1975
emission-control era), to sulphur from fuel impurities, and to
high-temperature sintering from rich-air-fuel-mixture operating
events.

\section{Failure: runaway reactions and the loss of cooling}\pathtag{core}

A reactor in which the heat-generation rate exceeds the heat-removal
rate is in \engterm{thermal runaway}. The temperature rises, the
reaction rate rises with it (Arrhenius), the heat generation rises
with the rate, and the temperature rises faster. The runaway
terminates only when the reactants are exhausted, the containment
fails, or the reactor is dismantled by an explosion or a venting
event. The failure mode is the most consequential in process
engineering and the most often realised in practice.

\engterm{The bifurcation diagram}. For a continuous stirred-tank
reactor with a single exothermic reaction cooled by a coolant at
temperature $T_c$, the heat-balance equation at steady state is
\[
U A (T - T_c) = (-\Delta H_r) V k(T) c^n,
\]
with $U$ the overall heat-transfer coefficient, $A$ the heat-transfer
area, $V$ the reactor volume, $c$ the reactant concentration, and
$n$ the reaction order. The left side is the heat-removal rate (linear
in $T$); the right is the heat-generation rate (exponential in $T$
through Arrhenius). The two curves may intersect at one, two, or
three steady states; when there are three, the middle one is
unstable, and the reactor can transition from the low-conversion
stable state to the high-conversion stable state with a small
perturbation. The high-conversion state often has temperature too
high for the reactor's mechanical containment.

\engterm{Adiabatic temperature rise}. The maximum temperature the
reactor could reach if no heat is removed and the reaction goes to
completion is
\[
\Delta T_{\text{ad}} = \frac{(-\Delta H_r) c_0}{\rho c_p},
\]
with $c_0$ the initial reactant concentration, $\rho$ the mixture
density, and $c_p$ its specific heat. For a typical organic exothermic
reaction with $\Delta H_r = -200\,\text{kJ}/\text{mol}$, $c_0 = 5\,
\text{mol}/\text{L}$, $\rho = 1000\,\text{kg}/\text{m}^3$, $c_p =
2\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$, the adiabatic temperature
rise is $500\,\si{\kelvin}$. The reactor at room temperature would,
on loss of cooling, rise to $800\,\si{\kelvin}$, a regime in which
the reactor's mechanical integrity is in doubt, secondary reactions
become rapid, and the contents may decompose or autoignite.

\engterm{The Bhopal mechanism}. On the night of 2--3 December 1984,
water entered a storage tank holding $\sim 42\,\text{tonnes}$ of
methyl isocyanate (MIC) at a Union Carbide pesticide plant in Bhopal,
India \cite{acc:icmr-bhopal-2004}. The water reacted exothermically
with the MIC; the heat raised the temperature; the rising temperature
accelerated both the water-MIC reaction and the MIC self-trimerisation;
a runaway ensued. The relief valve opened and discharged approximately
$30$ to $40\,\text{tonnes}$ of MIC vapour into the surrounding
neighbourhood. The flare tower was inoperable, the refrigeration
system for the storage tank had been disabled, the scrubber was
undersized for the discharge rate, and the alarm system was
ineffective in alerting the surrounding population. Thousands died in
the immediate release and tens of thousands suffered chronic injury
\cite{acc:icmr-bhopal-2004}. The technical mechanism is a
textbook example of thermal runaway whose magnitude was set by the
quantity stored (the safety culture of the era allowed bulk storage
of an extremely reactive intermediate), and whose consequence was set
by the failure of every mitigation layer.

\engterm{The runaway-failure pattern}. Process-safety analyses across
the chemical and petrochemical industries identify a common pattern
in major runaway-reaction releases: large stored inventory of an
exothermic reactant or product, loss of cooling or refrigeration,
incomplete relief-system sizing or routing, failed scrubbers or
flares, and absent or ineffective community-protection systems. The
2005 explosion at a Texas refinery isomerisation unit, killing fifteen
contractor workers in nearby trailers, exhibited a related class of
failure: a process unit was overfilled, liquid hydrocarbon overflowed
from the relief tower into a blowdown drum that vented to atmosphere
rather than to a flare, the resulting vapour cloud found an ignition
source, and the deflagration killed workers in nearby trailers. The
mechanism was loss of level control rather than loss of cooling, but
the family resemblance to the runaway pattern is exact: a reactive
process exceeded its containment because mitigation layers were
absent or non-functional.

\engterm{Mitigation hierarchy}. The standard hierarchy of process-safety
mitigations against runaway, in decreasing order of robustness:

\begin{enumerate}[leftmargin=*]
\item \engterm{Inherent safety}: reduce the inventory of hazardous
  material, dilute the reactants, choose a less energetic chemistry,
  run the reaction in continuous mode at low instantaneous inventory
  rather than batch mode with large hold-up.
\item \engterm{Passive safety}: containment vessels sized for the
  worst-case adiabatic temperature rise and pressure, vent paths to
  safe terminations, pressure-relief discs and valves.
\item \engterm{Active safety}: cooling-water systems, redundant
  refrigeration, automatic shutdown systems triggered by temperature
  or pressure excursions, interlocks that prevent the addition of
  reactant once a temperature threshold is exceeded.
\item \engterm{Procedural safety}: operator training, written
  procedures, alarm response protocols, emergency-response plans.
\end{enumerate}

The hierarchy is the practical content of the published process-safety
guidelines of the Center for Chemical Process Safety and of the
equivalent European regulations under the Seveso III Directive
\cite[ch.~26]{text:atkins-physical-chemistry}. The engineering
discipline of process safety is in moving each mitigation layer up
the hierarchy as far as it can go: inherent safety beats passive
safety beats active safety beats procedural safety. A plant whose
safety case relies on procedural responses to active failures of
passive containment of inherently unsafe chemistry is one in which the
failure of any mitigation layer initiates the next-deeper response,
each of which has a finite reliability.

\begin{principle}[Heat balance gates the runaway]
A reactor's stability against thermal runaway is gated by the local
slope of the heat-generation curve versus the heat-removal curve at
the operating point. When the heat-removal slope exceeds the
heat-generation slope, the reactor returns to the operating point
under perturbation; when the heat-generation slope exceeds the
heat-removal slope, the reactor runs away. The engineering content of
process-safety analysis is the construction of the heat-balance curve
for the chosen operating point, the identification of the
heat-removal margin, and the design of mitigation that preserves the
margin under every plausible failure scenario.
\end{principle}

\begin{estimation}
\textbf{Question.} Estimate the adiabatic temperature rise of a
$2\,\text{L}$ batch of $30\,\%$ hydrogen peroxide on uncontrolled
decomposition to water and oxygen.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The reaction $2\,\text{H}_2\text{O}_2 \to 2\,\text{H}_2\text{O} +
\text{O}_2$ has $\Delta H_r = -98\,\text{kJ}/\text{mol H}_2\text{O}_2$.
A $30\,\%$ solution has H$_2$O$_2$ mass fraction $0.30$, density
$\rho \approx 1.11\,\text{g}/\text{mL} = 1110\,\text{kg}/\text{m}^3$,
so the mass of H$_2$O$_2$ per litre is $0.30 \times 1110 = 333\,\text{g}$,
or $9.79\,\text{mol}/\text{L}$.

The released heat per litre is $9.79 \times 98 = 960\,\text{kJ}/\text{L}$.
The mixture's heat capacity is approximately $c_p \approx 4\,\text{kJ}/
(\text{kg}\cdot\si{\kelvin})$ (water-dominated). The temperature rise
in the adiabatic limit, ignoring phase change, is
\[
\Delta T_{\text{ad}} = \frac{960 \times 10^3\,\text{J}/\text{L}}{(1.11
\times 4000)\,\text{J}/(\text{L}\cdot\si{\kelvin})} \approx 216\,
\si{\kelvin}.
\]

The mixture starts at $\sim 25\,\degC = 298\,\si{\kelvin}$ and would
reach $\sim 514\,\si{\kelvin} = 241\,\degC$, except that the boiling
point of water at $1\,\text{atm}$ is $100\,\degC$. Long before the
adiabatic ceiling, the released heat boils the water, producing a
steam-and-oxygen mixture at high pressure if the container is sealed.

\smallskip
\textit{Postmortem.} The estimate captures the magnitude of the
hazard. The actual scenario is a phase-change problem rather than a
pure temperature-rise problem; the latent heat of vaporisation of
water ($2257\,\text{kJ}/\text{kg}$) absorbs $\sim 2500\,\text{kJ}/
\text{L}$ in evaporating the entire water content, more than the
$960\,\text{kJ}/\text{L}$ released. The actual outcome of an
uncontrolled decomposition is a mixture of warm water and oxygen at
elevated pressure, with the pressure set by the oxygen released
($9.79/2 = 4.9\,\text{mol O}_2/\text{L}$) at the resulting
temperature; the pressure can reach hundreds of bar in a sealed
container at room temperature. The engineering response is to vent
the container; commercial $30\,\%$ peroxide bottles include vented
caps for exactly this reason. The chapter's failure-section
machinery (adiabatic rise plus relief design) gives the right answer
on first attempt.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Titrate a household acid; quantify uncertainty]
The reader performs an acid-base titration on a household acid
(vinegar, lemon juice, or commercial citric acid) with a household
base (baking soda solution or ammonia solution), measures the
endpoint by pH paper or by colour-change indicator, and reports the
acid concentration with a propagated uncertainty.

\textbf{Track}. Household analytical chemistry.
\textbf{Hazard class}: low. Standard kitchen-scale concentrations of
acetic acid (vinegar, $\sim 5\,\%$ by weight), citric acid, sodium
bicarbonate, and household ammonia (dilute). No concentrated reagents.

\textbf{Procedure}.

\begin{enumerate}[leftmargin=*]
\item Standardise the base. Dissolve a known mass of sodium
  bicarbonate ($1.000\,\text{g}$, weighed on a kitchen scale with
  $0.01\,\text{g}$ precision) in $100\,\text{mL}$ of water. Compute
  the molar concentration ($1.000/84.0/0.100 = 0.119\,\text{mol}/
  \text{L}$) and the relative uncertainty in the concentration
  (mass uncertainty plus volume uncertainty).
\item Prepare a sample of the acid. Pipette $10.0\,\text{mL}$ of
  white vinegar (assayed at $\sim 5\,\%$ acetic acid by weight,
  density $\sim 1.005\,\text{g}/\text{mL}$) into a glass, with the
  volume measured by a kitchen measuring spoon (uncertainty $\pm 0.5\,
  \text{mL}$).
\item Add a few drops of an indicator (red cabbage juice is a
  practical kitchen indicator with colour change near pH $7$).
\item Add the standardised bicarbonate solution incrementally with a
  measuring cup or dropper, stirring after each addition, until the
  indicator colour changes. Record the volume added.
\item Compute the moles of bicarbonate consumed (equal to the moles of
  acid in the sample, by the $1{:}1$ stoichiometry of acetic acid
  with bicarbonate to form sodium acetate, water, and CO$_2$).
\item Compute the acid concentration in the sample (moles divided by
  sample volume) and convert to mass fraction.
\item Propagate the uncertainty: relative uncertainties in mass,
  volume, and endpoint volume add in quadrature. Report the result
  as a central value with a $1\sigma$ uncertainty.
\item Repeat the titration twice more and compare the three results.
  The standard deviation of the three runs is an empirical check on
  the propagated uncertainty.
\end{enumerate}

\textbf{Deliverable}. A three- to five-page lab report with the data
table, the uncertainty calculation, the comparison of measured to
label value, and a short discussion of which uncertainty source
dominated.

\textbf{Time}. Two hours of bench work, one to two hours of write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Balance the combustion of octane: $\text{C}_8\text{H}_{18} +
\text{O}_2 \to \text{CO}_2 + \text{H}_2\text{O}$. Compute the mass of
oxygen required to combust $1\,\text{kg}$ of octane.
\end{exercise}

\begin{exercise}
The rate constant of a first-order decomposition is $k = 0.020\,\si{\per\minute}$
at $25\,\degC$. Compute the half-life and the time required for $90\,\%$
conversion.
\end{exercise}

\begin{exercise}
A reaction has activation energy $E_a = 80\,\text{kJ}/\text{mol}$. Use
the Arrhenius equation to compute the ratio $k(50\,\degC)/k(25\,\degC)$.
\end{exercise}

\begin{exercise}
For the gas-phase equilibrium $\text{N}_2\text{O}_4 \rightleftharpoons
2\,\text{NO}_2$ at $25\,\degC$, $K_p = 0.143$. Compute the partial
pressure of NO$_2$ in equilibrium with N$_2$O$_4$ at $1.0\,\text{atm}$
total pressure, assuming pure N$_2$O$_4$ was charged initially.
\end{exercise}

\begin{exercise}
A buffer is prepared by mixing $0.10\,\text{mol}$ of acetic acid
($\text{p}K_a = 4.76$) with $0.10\,\text{mol}$ of sodium acetate in
$1.0\,\text{L}$ of water. Compute the pH. Compute the pH after the
addition of $0.010\,\text{mol}$ of HCl.
\end{exercise}

\begin{exercise}
The solubility product of silver chloride at $25\,\degC$ is $K_{sp} =
1.8 \times 10^{-10}$. Compute the molar solubility of AgCl in pure
water and in a $0.10\,\text{M}$ NaCl solution.
\end{exercise}

\begin{exercise}
A reaction with $\Delta H^\circ = -100\,\text{kJ}/\text{mol}$ and
$\Delta S^\circ = -120\,\text{J}/(\text{mol}\cdot\si{\kelvin})$ has
what equilibrium constant at $25\,\degC$ and at $300\,\degC$? Identify
the temperature above which the reaction becomes non-spontaneous.
\end{exercise}

\begin{exercise}
Compute the adiabatic temperature rise of a $1\,\text{L}$ batch of
$10\,\%$ aqueous ammonium nitrate undergoing decomposition, given
$\Delta H_r = -350\,\text{kJ}/\text{mol}$, density $\rho = 1.04\,
\text{g}/\text{mL}$, and specific heat $c_p = 3.7\,\text{kJ}/
(\text{kg}\cdot\si{\kelvin})$.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the integrated rate law for a first-order reaction in a batch
reactor at constant volume, starting from $-dc/dt = k c$.
\end{exercise}

\begin{exercise}
Derive the Henderson-Hasselbalch equation $\text{pH} = \text{p}K_a +
\log_{10}([\text{A}^-]/[\text{HA}])$ from the acid-dissociation
equilibrium expression.
\end{exercise}

\begin{exercise}
Starting from $\Delta G = \Delta G^\circ + RT \ln Q$ and the
equilibrium condition $\Delta G = 0$, derive $\Delta G^\circ = -RT
\ln K$.
\end{exercise}

\begin{exercise}
Derive the van't Hoff equation $d \ln K/dT = \Delta H^\circ/(RT^2)$
from the Gibbs-Helmholtz relation $d(\Delta G^\circ/T)/dT = -\Delta
H^\circ/T^2$.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the pH of a $0.1\,\text{M}$ solution of acetic acid ($K_a =
1.8 \times 10^{-5}$) without solving the full quadratic.
\end{exercise}

\begin{exercise}
Estimate the rate constant ratio $k(80\,\degC)/k(20\,\degC)$ for a
reaction whose half-life at $20\,\degC$ is one hour and whose
activation energy is $50\,\text{kJ}/\text{mol}$.
\end{exercise}

\begin{exercise}
Estimate the equilibrium yield of ammonia in the Haber process at
$500\,\degC$ and $200\,\text{bar}$ from $K_p$ data, assuming
stoichiometric feed.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a continuous stirred-tank reactor model for a single
exothermic reaction with Arrhenius kinetics and a coolant jacket.
Plot the steady-state heat-generation and heat-removal curves as a
function of reactor temperature and identify the multiple steady
states. Verify that the middle steady state is unstable by simulating
small perturbations.
\end{exercise}

\begin{exercise}
Implement a stoichiometric balance for a multi-reaction network with
five reactions and ten species. Given inlet flows and conversion of
each reaction, compute the outlet composition. Verify against a hand
calculation for the methane steam-reforming network.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify the cooling capacity required for a batch reactor running an
exothermic reaction with $\Delta H_r = -150\,\text{kJ}/\text{mol}$,
reactant feed $5\,\text{mol}/\text{L}$, volume $1\,\text{m}^3$, and
desired isothermal operation at $40\,\degC$. State the assumed
reaction time and the worst-case heat-removal requirement.
\end{exercise}

\begin{exercise}
Design a buffer for a process operation that must hold pH at $7.4
\pm 0.1$ during addition of a strong acid. Identify a suitable
weak-acid-conjugate-base pair and compute the required concentrations.
\end{exercise}

\begin{exercise}
Specify the pressure-relief vent area for a $5\,\text{m}^3$ batch
reactor running an exothermic reaction whose adiabatic temperature
rise is $200\,\si{\kelvin}$ and whose maximum operating pressure is
$10\,\text{bar}$. Use a simplified ideal-gas approximation for the
generated vapour and state the relief-philosophy assumption.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A reaction is reported to be first-order, with rate constants
measured at three temperatures: $k(25\,\degC) = 0.010\,
\si{\per\minute}$, $k(35\,\degC) = 0.022\,\si{\per\minute}$,
$k(45\,\degC) = 0.045\,\si{\per\minute}$. Extract the activation
energy and the pre-exponential factor from an Arrhenius plot.
Identify whether the data are consistent with a single-step
mechanism.
\end{exercise}

\begin{exercise}
A vendor's data sheet for a catalyst gives ``$95\,\%$ conversion at
$300\,\degC$'' without identifying the inlet concentration or the
contact time. Identify the missing information needed to assess
whether the catalyst meets a given process specification.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A batch reactor running an exothermic reaction at $80\,\degC$ in a
coolant jacket at $20\,\degC$ experiences a coolant-flow stoppage.
The reaction has $\Delta H_r = -100\,\text{kJ}/\text{mol}$, activation
energy $90\,\text{kJ}/\text{mol}$, and reactant concentration $3\,
\text{mol}/\text{L}$. Estimate the time to thermal runaway, defined as
the time when the temperature exceeds $200\,\degC$. Identify the
single mitigation that would have prevented runaway.
\end{exercise}

\begin{exercise}
A storage vessel for an exothermically self-reactive intermediate
loses its refrigeration. The released heat is initially absorbed by
the vessel's mass (steel, $c_p = 0.5\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$, mass $5000\,\text{kg}$) and by the intermediate
itself ($1000\,\text{kg}$, $c_p = 2\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$). Estimate the temperature rise after one hour of
adiabatic heat-up at a heat-generation rate of $5\,\text{kW}$,
assuming the kinetics do not accelerate appreciably over the
interval. Identify the two assumptions that fail at higher
temperature.
\end{exercise}

\begin{exercise}
A scrubber installed to absorb a vapour release from a pressure
relief is sized for a release rate of $1\,\text{kg}/\si{\second}$; the
relief actually delivers $50\,\text{kg}/\si{\second}$. Identify the
consequence at the scrubber outlet and the design philosophy that
should have prevented the mis-sizing.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A colleague proposes to increase the yield of an exothermic
equilibrium reaction by increasing the temperature. Identify the
trade-off the colleague is missing and state the correct strategy
for shifting the equilibrium toward products.
\end{exercise}

\begin{exercise}
A vendor claims that a new catalyst will allow ammonia synthesis at
$100\,\degC$ and $1\,\text{atm}$ rather than at the Haber-Bosch
conditions. Write a one-paragraph reply identifying the thermodynamic
constraint the vendor would need to confront and the experimental
test that would settle the claim.
\end{exercise}

\begin{exercise}
A student argues that a reaction with $\Delta G^\circ > 0$ cannot
occur. Identify the qualification this claim requires (the relation
between the standard free energy and the actual driving force at
non-standard composition) and state when the claim is valid.
\end{exercise}
