\chapter{Planetary boundaries and biosphere engineering}
\label{vol12:ch12}

\epigraph{We have engineered ourselves out of the conditions that produced us.}{Author, after Crutzen}

\chapmeta{Half-life: medium-life. Archetypes: scaling, balance, ethics, failure. Exercise target: 20. Project track: simulation (assess one planetary boundary; identify the engineering levers). Hazard class: high (planetary stakes; some boundaries already crossed). Reader path default: standard, with the planetary-boundaries-framework and biogeochemical-cycles sections core and the polar-regions section mastery. Named cases: Newfoundland cod-fishery collapse 1992 (\cite{paper:myers-cod-2007}); the Atlantic-meridional-overturning circulation as boundary near collapse (\cite{paper:rahmstorf-amoc-2015}); the nitrogen and phosphorus boundaries as crossed (\cite{paper:rockstrom-2009}).}

\noindent
The planetary-boundaries framework (Rockstr{\"o}m et al., 2009; \cite{paper:rockstrom-2009}; updated in Steffen et al., 2015; \cite{paper:steffen-2015}; latest in Richardson et al., 2023; \cite{paper:richardson-2023}) identifies $9$ Earth-system processes whose contemporary state determines whether the planet remains in a Holocene-like state suitable for civilization, or transitions to an Anthropocene state whose human-suitability is uncertain. As of the 2023 assessment, $6$ of the $9$ boundaries have been crossed (climate change; biosphere integrity; biogeochemical flows of nitrogen and phosphorus; land-system change; freshwater change; novel entities including synthetic chemicals); $3$ remain within the safe zone (stratospheric ozone, after the Montreal Protocol-driven recovery; ocean acidification, currently approaching the boundary; aerosol loading, regionally variable). The chapter takes the engineering implication of the framework as the topic.

\section{The planetary-boundaries framework}\pathtag{core}

The planetary-boundaries framework synthesises the Earth-system science into a small number of quantified thresholds whose crossing carries substantial probability of large-scale, abrupt, or irreversible change. Each boundary has a quantified safe-operating limit, a current-state assessment, and a high-risk zone beyond which transition becomes likely. The framework is not the totality of Earth-system science but is a useful integration for policy-and-engineering purposes; it has been criticised for the specific quantifications and for the assumption that the boundaries are independent (most are coupled), and it has been refined across the three published assessments.

The engineering interpretation: each boundary is a constraint on the cumulative effect of human activity that determines whether the planet's bio-physical systems remain in a state compatible with civilization. The engineering profession's contribution: the specific intervention design for keeping each boundary within or returning each crossed boundary to the safe zone. The engineering profession's responsibility: the recognition that the boundaries exist, the documentation of where current activities push them, the engagement with the institutional structures that determine whether the responses are adequate.

\begin{archetype}[Scaling: planetary metabolism]
The planet's metabolism: $\approx 120$ GtC/yr of net primary production by terrestrial ecosystems; $\approx 50$ GtC/yr by marine ecosystems; $\approx 100$ Mt of nitrogen fixed by lightning, plus $\approx 100$ Mt of biological nitrogen fixation, plus $\approx 120$ Mt of human industrial nitrogen fixation (Haber-Bosch); $\approx 20$ Mt of phosphorus mobilised from rock weathering, plus $\approx 20$ Mt of mined phosphorus added to agricultural soils. Human activity has roughly doubled the active nitrogen and phosphorus cycles, with substantial consequences for downstream ecosystems. The engineering-and-scale insight: human activity now operates at the same magnitude as the major bio-geochemical cycles, and human decisions about energy, agriculture, and industry directly affect the planet's metabolism.
\end{archetype}

\begin{estimation}
\textbf{Question.} Estimate the fraction of global net primary production currently appropriated by human activity (HANPP) and identify the implications for biosphere integrity.

\textbf{Reasoning.} Global NPP: $\approx 60$ GtC/yr terrestrial. Human appropriation through agriculture (crops, livestock fodder, harvested wood, fuel), through land-use change (the reduction in NPP from converting natural ecosystems to less-productive agricultural ones), and through land-use intensity: $\approx 25\%$ of potential NPP per Haberl et al. (2007), with more recent estimates in the $25$-$40\%$ range. Conclusion: roughly a quarter of the planet's terrestrial photosynthetic output is human-appropriated; the remainder supports the wild biosphere. The implication for biosphere integrity: any further increase in HANPP comes at the cost of the wild biosphere; the planetary-boundaries framework's biosphere-integrity boundary identifies this trade-off.
\end{estimation}

\section{Biogeochemical cycles as engineered domain}\pathtag{core}

The major biogeochemical cycles: carbon (atmosphere; ocean; biosphere; soil; rocks; fossil fuels); nitrogen (atmosphere; soil; biosphere; water; agricultural; industrial); phosphorus (rocks; soil; biosphere; water; agricultural; industrial); sulfur (atmosphere; ocean; biosphere; combustion); water (atmosphere; oceans; rivers; lakes; groundwater; glaciers); oxygen (atmosphere; ocean; biosphere). Each cycle has natural rates and reservoirs; human activity has perturbed each to varying degrees.

The carbon cycle perturbation is the climate-change driver discussed in Chapter 11. The nitrogen and phosphorus cycle perturbations are the eutrophication driver: the synthetic fertiliser produced by Haber-Bosch (Volume XII Chapter 5) and the mined phosphate fertiliser both produce nutrient pollution that has consequences for freshwater and coastal-marine ecosystems. The chapter's failure section discusses the Gulf of Mexico dead zone (driven by Mississippi-basin nitrogen and phosphorus runoff) and the Baltic Sea hypoxia (driven by surrounding-region nutrient input) as the contemporary cases.

\begin{principle}[Biogeochemical cycles as constraint]
The biogeochemical cycles are constraints on the cumulative scale at which human activity can operate without destabilising the bio-physical systems that civilization depends on. The engineering response: explicit accounting of the cumulative flux of each cycle's anthropogenic perturbation; the technology and management interventions that reduce the perturbation to within sustainable rates; the engagement with the agricultural, industrial, and energy systems that produce the perturbations. The engineering and policy interface is the agricultural-management, industrial-process, and energy-system decisions that determine the cumulative flux.
\end{principle}

\section{Biodiversity and ecosystem services}\pathtag{standard}

Biodiversity is the variety of life at gene, species, and ecosystem levels. The contemporary state: extinction rates substantially elevated above background ($\sim 100$-$1{,}000\times$ background per most estimates; IPBES 2019); some major groups (amphibians, reef-building corals) at high risk; some ecosystems (tropical forests, coral reefs, freshwater systems) at substantial pressure; population sizes of monitored vertebrates declining by approximately $70\%$ since 1970 per WWF's Living Planet Index. The drivers: land-use change (deforestation, agricultural expansion), direct exploitation (fisheries, hunting, trade), climate change, pollution, invasive species. The Intergovernmental Science-Policy Platform on Biodiversity and Ecosystem Services (IPBES) provides the analogous role to the IPCC for the biodiversity-and-ecosystem domain.

Ecosystem services are the benefits humans derive from ecosystems: provisioning (food, fibre, water, fuel); regulating (climate regulation, water purification, pollination, pest control); cultural (recreation, aesthetic, spiritual); supporting (soil formation, nutrient cycling). The economic valuation of ecosystem services (Costanza et al., 1997 and successors) is in the trillions of dollars annually at the global scale; the policy implication is that the maintenance of ecosystem services is an engineering objective comparable in scale to the build-and-maintain engineering of the previous chapters.

\section{Soil at planetary scale}\pathtag{standard}

Soil is the thin layer (typically $0.1$-$1$ m thick) where rock, water, atmosphere, and biosphere meet; it is the substrate of nearly all terrestrial agriculture, the storage reservoir for substantial atmospheric carbon (soil organic carbon at $\sim 1{,}500$-$2{,}500$ GtC, comparable to the atmospheric carbon pool), and a key component of the planetary biogeochemical cycles. The contemporary state: substantial soil degradation across the agricultural and pastoral lands, with the FAO estimating that $\sim 33\%$ of global soils are moderately to highly degraded (current as of 2024). The drivers: erosion (water and wind, accelerated by intensive agriculture); salinisation (irrigation in arid areas with inadequate drainage); compaction (heavy agricultural machinery); contamination (industrial and pesticide); organic-matter loss (intensive cultivation; insufficient organic-matter return).

The engineering response: conservation tillage; cover cropping; integrated nutrient management; soil-organic-matter restoration; appropriate irrigation-and-drainage management; remediation of contaminated soils. The engineering and the political-economic context interact: most soil-conservation practices have private benefits and public benefits that may not align in time or actor, and the institutional structures (extension services; subsidy regimes; tenure security; carbon markets) determine whether the practices are adopted. The Vol XII Chapter 5 agricultural-engineering case is the parallel discussion at the field-and-farm scale; this chapter takes the global-soil-system as the unit.

\section{Oceans}\pathtag{standard}

The oceans are the planet's dominant heat and carbon sink: they have absorbed $\sim 90\%$ of the heat from anthropogenic forcing and $\sim 30\%$ of the cumulative anthropogenic CO$_2$. The consequences: warming (with substantial regional variation; the upper-ocean warming most consequential for marine ecosystems); acidification (CO$_2$ absorption produces carbonic acid; pH has dropped $\sim 0.1$ unit since pre-industrial, a $\sim 30\%$ increase in hydrogen-ion concentration); deoxygenation (the warming-and-stratification-driven decrease in oxygen content; substantial in some regions including the Pacific oxygen-minimum zones); rising sea level ($\sim 22$ cm since 1880, with the rate accelerating; projected to reach $\sim 30$-$100$ cm by 2100 depending on emissions trajectory and ice-sheet dynamics).

The fisheries pressure adds to the climate-driven stresses. Approximately $35\%$ of monitored fish stocks are over-fished, with regional and species variation; the most-pressed fisheries (Atlantic cod; bluefin tuna; some shark species) are at small fractions of historical biomass. The Newfoundland cod-fishery collapse \cite{paper:myers-cod-2007} of 1992 is the iconic case: a moratorium imposed only after the stock had collapsed to less than $1\%$ of historical levels; the stock has not substantially recovered $30$ years later despite the moratorium continuing; the regional economy and the cod-dependent communities have not recovered either. The engineering lesson: fisheries-management decisions that ignore the engineering analysis of stock-and-fleet dynamics produce predictable failures that the engineering profession (in the contemporary sense, including the fisheries-science and marine-resource-management communities) is responsible for surfacing.

\section{Polar regions and cryosphere}\pathtag{mastery}

The polar regions are the most rapidly changing parts of the contemporary climate system. Arctic warming has been roughly $4\times$ the global mean rate (the Arctic-amplification phenomenon); Arctic September sea-ice extent has declined $\sim 50\%$ since 1980; Greenland and Antarctic ice sheets have accelerated their mass loss (Greenland $\sim 280$ Gt/yr; Antarctica $\sim 150$ Gt/yr; combined contributing $\sim 1$ mm/yr to global sea-level rise; current as of 2024). The cryosphere changes have substantial regional and global consequences: sea-level rise, ecosystem disruption, indigenous-community displacement, ocean-circulation changes including the Atlantic Meridional Overturning Circulation (AMOC) under stress (\cite{paper:rahmstorf-amoc-2015}), and the methane-release potential from thawing permafrost (a substantial climate-feedback uncertainty).

The Arctic-Antarctic engineering frontier is the early-stage discussion of geoengineering options (refreezing the Arctic; supporting the West Antarctic Ice Sheet; ocean-current modification) that are at the technical-feasibility and ethical-governance stages. The engineering profession's role: the credible technical analysis of these options; the explicit acknowledgement of the planetary-scale risks; the refusal to advance the deployment in the absence of credible governance frameworks.

\section{Engineering for biosphere health}\pathtag{standard}

Engineering for biosphere health is the application of engineering disciplines to the maintenance and restoration of biosphere integrity. The scope: ecosystem-restoration engineering (the restoration of damaged ecosystems to functioning condition; large-scale examples include the European LIFE-funded restoration projects, the US Comprehensive Everglades Restoration Plan, the Chinese Loess Plateau restoration); pollution-reduction engineering at the source (the cleaner-production approach that reduces effluent and waste before they enter the environment); invasive-species management (biological-control engineering; eradication programmes); habitat-corridor engineering (wildlife crossings; landscape connectivity); urban-ecology engineering (the integration of biodiversity into urban infrastructure).

The engineering profession's collaboration with the ecology and conservation-biology communities is one of the contemporary developments; the integration is not always smooth but is producing increasing alignment around common engineering-and-ecology methods. The IPBES, the UN Convention on Biological Diversity, the various regional conventions (Ramsar for wetlands; Bonn Convention for migratory species) are the institutional infrastructure.

\section{Failure: nutrient-cycle failures, fisheries collapse, ozone, PFAS}\pathtag{core}

The chapter's failure cases illustrate the spectrum of planetary-boundary-related engineering outcomes.

The Newfoundland cod-fishery collapse \cite{paper:myers-cod-2007} is the case of fisheries-management failure: stock-assessment errors compounded by fleet-overcapacity, weak regulatory enforcement, and political reluctance to impose the catch limits that the science indicated; the eventual moratorium came after the stock had collapsed; the recovery has been substantially incomplete in $30$ years of zero-or-low fishing. The engineering interpretation: the failure was not principally one of marine-biology understanding but of the integration of biological knowledge with fleet-management engineering and with the political-economic institutional context.

The nutrient-cycle perturbation (nitrogen and phosphorus, both boundary-crossed per the planetary-boundaries assessment) is the case of biogeochemical-engineering failure: the Haber-Bosch process and the phosphate-fertiliser industry deliver the nutrients that feed the global population but at the cost of substantial downstream consequences (eutrophication of freshwater and coastal systems; groundwater contamination; greenhouse-gas emissions from soil). The engineering response (nutrient-use-efficiency improvements; integrated nutrient management; downstream wetland-and-buffer engineering) is well-developed at the field-and-farm scale; the implementation at the global-cumulative-flux scale is the binding constraint.

The Montreal Protocol \cite{law:montreal-protocol-1987} is the case of planetary-boundary success: the CFC and HCFC phase-out has reduced atmospheric concentrations of these substances and is allowing the ozone layer to recover; the stratospheric-ozone boundary has moved from crossed to within the safe zone. The engineering interpretation: the success is the result of a combination of credible scientific assessment (the ozone-depletion-mechanism science of Molina and Rowland; the Antarctic-ozone-hole observation of Farman et al.), engineering response (the HFC and other refrigerant substitutes), and institutional design (the protocol's structure with universal participation, differentiated obligations, technology-transfer, and the Multilateral Fund). The Kigali Amendment (\cite{law:kigali-amendment-2016}) extends the regime to HFCs (which have climate-but-not-ozone consequences) and is one of the contemporary international-engineering-governance successes.

The PFAS (per-and-polyfluoroalkyl substances; the ``forever chemicals'') case discussed in Vol X Chapter 15 is the case of novel-entities boundary problem: a class of substances of $\sim 12{,}000$ individual chemicals with extreme environmental persistence, widespread contamination of water and biota, and emerging human-health-effects evidence; the regulatory response is at the early stages and is highly variable across jurisdictions. The engineering lesson: novel-entities introduction without adequate environmental-and-health-effects assessment produces consequences whose remediation is substantially more expensive than the prevention would have been. The Vol XII Chapter 15 ozone success is the counter-example: when the assessment is done credibly and the response is institutionally backed, the planetary-boundary excursion can be reversed.

\begin{principle}[Engineering for biosphere health as the cumulative obligation]
Engineering for biosphere health is the cumulative obligation of the engineering profession over the decades during which the contemporary planetary-boundary excursions are addressed. The discipline: credible accounting of cumulative human impact on the planetary biogeochemical and ecological systems; the design of interventions that reduce the impact within sustainable limits; the engagement with the agricultural, industrial, energy, and consumer systems whose cumulative activity determines the impact; the engagement with the institutional structures that determine whether the interventions are adopted. The engineering profession's failure to engage is itself a failure mode for which the profession bears responsibility.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Planetary-boundary engineering assessment]
Choose one of the planetary boundaries (climate; biosphere; nitrogen; phosphorus; land-use; freshwater; novel entities; ocean acidification; stratospheric ozone; atmospheric aerosols). Document the current state of the boundary (the quantified position relative to the safe zone); document the principal anthropogenic drivers (the sectors, activities, and decisions that determine the boundary's position); identify the engineering levers (the technical interventions that could shift the boundary toward the safe zone); estimate the deployment requirements (capital, workforce, timeline, institutional structures); identify the political-economic constraints. Produce an engineering-credible assessment of what would be required to return the boundary to the safe zone (if crossed) or to keep it there (if not).
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A river basin receives $50{,}000$ tonnes/yr of nitrogen runoff from agriculture, of which $30\%$ reaches the coastal zone; the receiving estuary has area $1{,}000\,\text{km}^2$. Compute the per-area nitrogen loading rate; compare to the hypoxia-threshold loading rate of $\sim 50\,\text{kg/ha/yr}$.
\end{exercise}

\begin{exercise}
A fishery stock has biomass $B = 200{,}000$ tonnes, intrinsic growth rate $r = 0.3$/yr, and is harvested at rate $h$. Compute the maximum sustainable yield and the corresponding harvest rate.
\end{exercise}

\begin{exercise}
A degraded forest landscape sequesters $5\,\text{tC/ha/yr}$ during regrowth over $30$ years; a restoration programme covers $100{,}000$ ha. Compute the cumulative carbon sequestration and the equivalent CO$_2$ tonnes.
\end{exercise}

\begin{exercise}
A novel chemical with $50$-year half-life is released at $1{,}000$ tonnes/yr. Compute the steady-state environmental stock and the time to reach $95\%$ of steady-state.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the relationship between agricultural nitrogen-use-efficiency and downstream nutrient loading. Identify the management improvements that reduce loading without yield loss.
\end{exercise}

\begin{exercise}
Derive the population-dynamics model for a fish stock under predation and harvest. Identify the conditions under which the stock collapses.
\end{exercise}

\begin{exercise}
Derive the ocean-acidification trajectory as a function of atmospheric CO$_2$ concentration; identify the threshold concentration for substantial coral-reef decline.
\end{exercise}

\begin{exercise}
Derive the cost-of-restoration estimate for a degraded ecosystem as a function of the degradation severity, the area, and the target restoration condition.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the global engineering and policy investment required to bring the nitrogen and phosphorus planetary boundaries back to safe-zone levels.
\end{exercise}

\begin{exercise}
Estimate the global capital required for the deployment of soil-conservation practices across the world's agricultural lands.
\end{exercise}

\begin{exercise}
Estimate the area of land required for ecosystem-restoration commitments comparable to the Bonn Challenge (350 million ha by 2030).
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate the recovery dynamics of a collapsed fishery under different harvest-control regimes. Identify the regimes that produce recovery versus those that produce continued collapse.
\end{exercise}

\begin{exercise}
Simulate the nitrogen-cascade through a watershed under different agricultural-management scenarios. Identify the management improvements that produce target loading reductions.
\end{exercise}

\begin{exercise}
Simulate the trajectory of a novel-chemical contamination through environmental media under different release-reduction policies. Identify the policy interventions that contain the contamination.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a wetland-and-buffer system for nutrient interception in an agricultural watershed of $50{,}000$ ha. Specify the wetland area, the buffer width, the placement, the cost, the nutrient-removal effectiveness.
\end{exercise}

\begin{exercise}
Design an ecosystem-restoration programme for a $10{,}000$-ha degraded landscape. Specify the restoration techniques, the financing, the monitoring, the institutional structure.
\end{exercise}

\begin{exercise}
Design a fisheries-management regime for a regional fishery currently under collapse risk. Specify the harvest limits, the enforcement mechanism, the institutional structure, the just-transition support for affected fishing communities.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Discuss the Newfoundland cod-fishery collapse as an engineering and institutional failure. Identify the parallels with the planetary-boundary excursions; identify the engineering interventions that would have prevented the collapse.
\end{exercise}

\begin{exercise}
Argue: the planetary-boundaries framework is the appropriate organising concept for engineering at civilization scale; the engineering profession should adopt it as a planning framework. Argue the opposite. Identify the engineering implications of each position.
\end{exercise}

\begin{exercise}
Discuss the Montreal Protocol as a planetary-boundary engineering success. Identify the engineering, scientific, and institutional factors; identify the differences with the climate-change case that limit the analogy.
\end{exercise}
