\chapter{Climate engineering, mitigation, adaptation}
\label{vol12:ch11}

\epigraph{The atmosphere is now an engineered artifact, whether engineers wish it or not. The discipline is to engineer it deliberately rather than incidentally.}{Author}

\chapmeta{Half-life: medium-life. Archetypes: scaling, ethics, control. Exercise target: 20. Project track: simulation (assess one mitigation pathway's engineering credibility). Hazard class: high (planetary stakes; multi-generational consequences). Reader path default: standard, with the mitigation and adaptation sections core and the SRM-governance section mastery. Named cases: Solar Radiation Management Governance Initiative (\cite{paper:srmgi-2011}); the Montreal Protocol as climate-engineering success (\cite{law:montreal-protocol-1987}).}

\noindent
The Earth's climate is the largest single engineered consequence of human activity. The cumulative anthropogenic CO$_2$ emissions since 1850 ($\approx 2{,}500$ GtCO$_2$; \cite{std:ipcc-ar6-wg1-2021}) have produced atmospheric CO$_2$ concentration of $\approx 422$ ppm (current as of 2024), warming of $\approx 1.2\degC$ above pre-industrial average, and the climate-system consequences the IPCC has documented in successive assessment reports. The future trajectory is the engineering-and-political question of the century: limiting warming to $1.5$-$2\degC$ by 2100, the current Paris-Agreement objective, requires the energy and industrial transformations described in earlier chapters of this volume plus, increasingly, deliberate engineering at the climate-system scale. The chapter takes the climate-system as the engineering object, distinguishes mitigation (reducing emissions), adaptation (managing consequences), carbon-dioxide removal (active removal of accumulated CO$_2$), and solar-radiation management (modifying the Earth's radiative balance directly), and treats each at the engineering and governance levels.

\section{Climate as engineered domain whether we want it or not}\pathtag{core}

The contemporary climate is engineered in the same sense that a polluted river is engineered: the human intervention has determined the state, even though the intervention was incidental rather than deliberate. Greenhouse-gas concentrations: CO$_2$ from $\approx 280$ ppm in 1850 to $\approx 422$ ppm in 2024; CH$_4$ from $\approx 700$ ppb to $\approx 1{,}930$ ppb; N$_2$O from $\approx 270$ ppb to $\approx 336$ ppb; the halocarbons (CFCs, HCFCs, HFCs) from zero to $\approx 1{,}100$ ppt of equivalent radiative forcing (with substantial reductions in CFCs from the Montreal Protocol regime; \cite{law:montreal-protocol-1987}). Total radiative forcing from these gases plus other anthropogenic contributors: $\approx 2.7\,\text{W/m}^2$ as of 2024, against the planet's $\approx 240\,\text{W/m}^2$ baseline solar absorption.

The engineering interpretation: the climate response is the integrated consequence of human energy, agriculture, industrial, and land-use decisions over $170$ years. The contemporary engineering question is whether the next $170$ years' decisions will continue the same pattern (the high-emissions scenarios; warming reaching $3$-$4\degC$ by 2100; substantial ecosystem and human-system disruption) or shift to a different pattern (the Paris-consistent scenarios; warming limited to $1.5$-$2\degC$; substantial but bounded disruption). The engineering profession contributes to whichever pattern emerges through its role in the energy, industrial, agricultural, transport, and buildings sectors.

\begin{archetype}[Scaling: planetary energy balance]
The Earth's radiative balance: $\approx 340\,\text{W/m}^2$ of incoming shortwave solar radiation, $\approx 100\,\text{W/m}^2$ reflected by clouds and surface (albedo $\approx 0.3$), $\approx 240\,\text{W/m}^2$ absorbed by atmosphere and surface, and at equilibrium $\approx 240\,\text{W/m}^2$ of outgoing longwave radiation. The greenhouse effect raises the surface temperature above the radiative equilibrium with the top-of-atmosphere by the absorption-and-re-emission of longwave radiation in the atmosphere. Anthropogenic greenhouse-gas forcing of $\approx 2.7\,\text{W/m}^2$ shifts the equilibrium toward higher surface temperatures; the climate-system response (with the partly-equilibrated transient response and the larger eventual equilibrium response, plus feedbacks from water vapour, ice-albedo, clouds, biogeochemistry) determines the warming per unit of forcing. The equilibrium climate sensitivity (warming at doubled CO$_2$) is $\approx 3\degC$ per the IPCC AR6 likely range of $2.5$-$4\degC$.
\end{archetype}

\begin{estimation}
\textbf{Question.} Estimate the remaining global carbon budget for limiting warming to $1.5\degC$ with $67\%$ probability and identify the implications for the global emissions trajectory.

\textbf{Reasoning.} IPCC AR6 estimates the remaining 1.5\degC carbon budget (from 2020) at $\approx 400$ GtCO$_2$ at $67\%$ probability \cite{std:ipcc-ar6-wg1-2021}. Current annual global emissions: $\approx 40$ GtCO$_2$/yr (current as of 2024). Time at current emissions: $\approx 10$ years remaining; by 2030 the budget would be exhausted at current rates. The implied global emissions trajectory: rapid reduction starting immediately to reach net-zero by $\approx 2050$ with declining emissions throughout the 2020s and 2030s. The engineering implication: the energy-system, industrial, agricultural, transport, and buildings transformations described in this volume must accelerate substantially from current rates to keep the $1.5\degC$ target achievable.
\end{estimation}

\section{Mitigation engineering: where the most leverage is}\pathtag{core}

Mitigation engineering is the reduction of greenhouse-gas emissions across the major sources. The sectoral allocation of contemporary emissions: energy supply $\approx 35\%$; industry $\approx 24\%$; transport $\approx 16\%$; buildings $\approx 6\%$; agriculture, forestry, land-use $\approx 19\%$ (IPCC AR6 WG3 data). Each sector has its own mitigation engineering portfolio, described in earlier chapters: energy in Vol XII Ch 2; industry in Vol VIII; transport in Vol XII Ch 4; buildings in Vol XII Ch 7; agriculture in Vol XII Ch 5. The integration is the cross-sectoral mitigation pathway, with the IPCC AR6 WG3 documenting $\sim 1{,}200$ scenarios consistent with $1.5\degC$ or $2\degC$ targets \cite{std:ipcc-ar6-wg3-2022}.

The mitigation-engineering frontier sectors (the hardest to decarbonise) are steel, cement, ammonia, plastics, long-haul aviation, long-haul shipping, and heavy long-haul transport. Each has its own decarbonisation pathway: steel via hydrogen-direct-reduced iron plus electric-arc-furnace; cement via low-carbon binders, carbon capture, and clinker substitution; ammonia via green hydrogen-fed Haber-Bosch; plastics via biomass-feedstock and recycling; aviation via sustainable aviation fuel and the longer-term hydrogen aircraft concepts; shipping via ammonia, methanol, or hydrogen fuels; heavy trucks via battery and hydrogen-fuel-cell systems. The technologies are at varying stages of demonstration and commercialisation; the engineering and policy challenge is the deployment timeline.

\begin{principle}[Mitigation as the highest-leverage engineering response]
For any given climate target, the highest-leverage engineering response is the mitigation of emissions, because the avoided-emission has multi-century atmospheric persistence and the deployed mitigation technology has multi-decade operational life. Carbon-dioxide removal and solar-radiation management address either the residual emissions that mitigation cannot reach or the overshoot of the trajectory that mitigation has not contained; neither is a substitute for the mitigation, and the contemporary policy debate that frames them as alternatives is conceptually mistaken. The engineering profession's contribution is the credible documentation of what mitigation can achieve, at what cost, on what timeline, with what technologies.
\end{principle}

\section{Adaptation engineering}\pathtag{core}

Adaptation engineering is the management of climate-change consequences across the engineered and natural systems. The principal adaptation domains: water (the design-basis adjustments for flooding, drought, snowpack changes; Vol XII Ch 3); agriculture (the crop-and-region adjustments to changing climate; Vol XII Ch 5); coastal-zone (sea-level rise; storm-surge; managed retreat); urban (heat-island mitigation; flood-control; building-design adaptation; Vol XII Ch 7); transport (infrastructure design-basis adjustments); health (heat-wave response; vector-borne disease; air-quality); ecosystems (species-and-habitat management). Each domain has its own adaptation engineering at different stages of development.

The adaptation-engineering finance gap is substantial: the UNEP Adaptation Gap Report estimates global adaptation finance needs at $\approx \$160$-$\$340$ billion/yr by 2030 in developing countries alone, against current financing of $\approx \$20$ billion/yr (current as of 2023; \cite{web:unep-adaptation-gap-2023}). The engineering profession's contribution: credible cost estimates for the specific adaptation interventions; credible effectiveness estimates; the integration of adaptation with development planning so that the adaptation investments serve broader development objectives. The just-adaptation considerations parallel the just-transition considerations of Vol XII Ch 2: the populations most exposed to climate change are typically the populations with the least adaptation resources, and the engineering plans must address this distribution explicitly.

\section{Carbon dioxide removal: the menu}\pathtag{standard}

Carbon-dioxide removal (CDR) is the active extraction of CO$_2$ from the atmosphere followed by its long-term storage. The technology portfolio includes: afforestation and reforestation (planting and re-establishing forests; $\sim \$10$-$\$50$/tCO$_2$; substantial land area; permanence-of-storage concerns from forest fires, harvest, climate stress); bioenergy with carbon capture and storage (BECCS; growing biomass, combusting it for energy, capturing the resulting CO$_2$ and injecting it underground; $\sim \$100$-$\$200$/tCO$_2$; substantial land-and-water resource requirements; deployment-scale concerns); direct air capture with geological storage (DACCS; chemical capture of CO$_2$ from ambient air, injection into geological reservoirs; $\sim \$200$-$\$600$/tCO$_2$ at current early-stage costs; technology demonstration at $\sim 10{,}000$ tCO$_2$/yr scale as of 2024); enhanced rock weathering (mineral surfaces that bind atmospheric CO$_2$ as carbonates; emerging); ocean-based methods (ocean alkalinity enhancement; ocean fertilisation; coastal blue-carbon; mixed evidence and substantial governance concerns); soil-carbon sequestration (agricultural-management changes that increase soil organic carbon; permanence and measurement challenges).

The IPCC's net-zero pathways universally include some CDR, with the range from $\sim 100$ MtCO$_2$/yr to $\sim 10$ GtCO$_2$/yr by 2050 depending on the scenario. The engineering judgement: the CDR menu has options that are technologically credible at the GtCO$_2$/yr scale required, but the deployment infrastructure, the financing structures, and the verification infrastructure are not yet at scale. The engineering profession's contribution is the credible documentation of what each CDR option can deliver, at what cost, with what side effects, on what timeline.

\section{Solar radiation management: governance and risk}\pathtag{mastery}

Solar-radiation management (SRM) is the deliberate modification of Earth's radiative balance to offset warming without removing the CO$_2$ that caused it. The leading concepts: stratospheric aerosol injection (SAI; introducing sulphate or other aerosols into the stratosphere to reflect sunlight; analogous to the natural effect of large volcanic eruptions such as Pinatubo 1991 which produced $\approx 0.5\degC$ of global cooling for $\sim 2$ years); marine cloud brightening (MCB; spraying sea-salt aerosols to enhance low-cloud reflectivity over selected regions); cirrus cloud thinning (reducing high-cloud greenhouse effect). The engineering feasibility of SAI is well-established (aircraft-based delivery; the costs at the $\sim 1\degC$ scale are estimated at $\$10$-$\$50$ billion/yr); the consequences are far less well-understood and are the subject of substantial scientific and ethical concern.

The Solar Radiation Management Governance Initiative \cite{paper:srmgi-2011} and subsequent governance work has framed the engineering question as inseparable from the governance question: any SRM deployment would have global consequences with distributional effects across regions; the decision to deploy or not is a planetary-scale decision that no single state or actor has the authority to make; the research, governance, and decision-making infrastructure required for credible SRM consideration is at the very early stages. The engineering profession's responsibility includes the credible technical analysis of SRM options, the explicit acknowledgement of the governance and ethical limitations, and the refusal to advance SRM deployment in the absence of the governance infrastructure.

\section{Climate engineering as international issue}\pathtag{standard}

Climate engineering is irreducibly international: the atmosphere is a global commons; emissions anywhere affect the climate everywhere; mitigation and adaptation investments are unevenly distributed across countries; the historical responsibility for emissions is concentrated in industrialised countries while the climate-impact exposure is concentrated in developing countries. The UNFCCC framework (the 1992 convention; the 1997 Kyoto Protocol; the 2015 Paris Agreement) is the international institutional infrastructure; the COP (Conference of Parties) annual meetings are the negotiation venue; the IPCC provides the scientific assessment.

The contemporary international engineering coordination includes: technology-transfer programmes (the developed-country obligation to support technology transfer to developing countries under the UNFCCC); the climate-finance commitments (the $\$100$ billion/yr commitment from developed to developing countries; with continuing implementation shortfalls); the loss-and-damage fund (established at COP27 in 2022; operational details still being negotiated); the carbon-border-adjustment mechanisms (EU CBAM operational from 2026; similar mechanisms under consideration in other jurisdictions). The engineering profession's contribution at the international level: the credible technical analysis that supports policy negotiation; the participation in international standards and verification regimes; the engagement with technology-transfer mechanisms.

\section{Failure: climate-policy failures and what engineers should learn}\pathtag{core}

The thirty-year contemporary climate-policy record has substantial mixed evidence on what works and what does not. The Montreal Protocol (\cite{law:montreal-protocol-1987}) is the case of climate-engineering success: the CFC phase-out has reduced atmospheric CFC concentrations and is allowing the ozone layer to recover; the regime's design (universal participation; differentiated obligations; finance for compliance; mandatory technology adoption; aggressive scope-expansion through subsequent amendments) is widely cited as the model for international environmental governance. The Kyoto Protocol (1997, in force 2005) is the case of partial success: it established the international compliance framework but the absence of the US and the absence of binding obligations on developing countries limited its impact. The Paris Agreement (2015) is the case of structural innovation: it replaced the top-down assigned-target framework with a bottom-up nationally-determined-contribution framework, traded compliance binding for political plausibility, and remains the operative regime; its effectiveness is contested and the gap between current commitments and the $1.5\degC$ trajectory is substantial.

The Vol X Chapter 15 climate-failure analysis described the lessons from the lead-petrol, ozone, and PFAS cases for climate engineering: surface the consequences before the harm is irreversible; protect the engineering and scientific judgement against industry pressure; design regulatory regimes for the actual rather than the assumed pathway. The engineering interpretation of the contemporary climate-policy record: the IPCC scientific assessment has been credible and the technical-engineering analysis has been credible; the political-economic execution has lagged the technical analysis; the engineering profession's responsibility is to maintain the credibility of the technical assessment, to be unsparing about what mitigation can and cannot achieve, and to engage with the political-economic context that determines whether the engineering plans become reality.

\begin{principle}[Climate engineering as the integrative test of the engineering profession's civilizational role]
Climate engineering is the integrative test of the engineering profession's civilizational role: it requires the technical literacy of all preceding volumes; the failure literacy of Volume X; the design discipline of Volume XI; the civilizational understanding of Volume XII. It requires the engineering profession's engagement with political-economic constraints that exceed the technical scope. It requires international coordination at a scale the engineering profession is only beginning to engage with. The engineering profession's response over the next decades will be the test of whether the civilizational understanding the project has tried to develop has been adequate to the responsibility the profession has been asked to bear.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Mitigation-pathway engineering credibility]
Choose a published mitigation pathway (an IPCC scenario; a national net-zero strategy; an industry decarbonisation roadmap; a corporate net-zero plan). Document the pathway's emissions trajectory, the assumed technology deployment, the assumed costs, the assumed land-and-resource requirements. Compare to the engineering reality: the current deployment rates, the current technology TRL, the current cost trajectories, the supply-chain constraints, the workforce constraints. Identify the points at which the pathway's assumptions exceed the demonstrated engineering capability; identify the additional investments, policy changes, or engineering breakthroughs required to make the pathway feasible. Identify what the pathway omits (the residual emissions; the adaptation needs; the just-transition support). Produce a credibility assessment that distinguishes the engineering-credible elements from the wishful elements.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A coal-fired power plant emits $0.85\,\text{tCO}_2/\text{MWh}$ and operates $7{,}000\,\text{h/yr}$ at $500\,\text{MW}$. Compute the annual CO$_2$ emissions in tonnes. Compute the gigatonnes-equivalent if the entire global coal fleet ($\sim 2{,}000\,\text{GW}$) operates at the same rate.
\end{exercise}

\begin{exercise}
A direct-air-capture plant captures $1$ MtCO$_2$/yr at $\$400$/tCO$_2$. Compute the annual cost and compare to $1\,\text{Mt CO}_2$/yr from a typical large coal-fired power plant; compute the cost-per-tCO$_2$ of avoided versus removed emissions.
\end{exercise}

\begin{exercise}
A tropical-reforestation programme plants $10{,}000\,\text{ha/yr}$ with carbon sequestration of $\approx 5\,\text{tC/ha/yr}$ over a $30$-year forest-growth period. Compute the annual carbon sequestration rate and the cumulative sequestration over $30$ years.
\end{exercise}

\begin{exercise}
A stratospheric aerosol injection programme aims for $1\degC$ of cooling effect. Estimate the annual aerosol deployment rate required (use the volcanic-analogy that Pinatubo's $\sim 17$ Mt of SO$_2$ produced $\sim 0.5\degC$ cooling for $\sim 2$ years).
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the relationship between cumulative emissions and warming (the TCRE; transient climate response to emissions) and identify the implications for the remaining carbon budget at a given temperature target.
\end{exercise}

\begin{exercise}
Derive the optimal allocation of mitigation effort across sectors given the marginal abatement-cost curves. Identify the implications for sector-by-sector decarbonisation pricing.
\end{exercise}

\begin{exercise}
Derive the radiative-forcing-to-temperature response function from the planetary energy balance. Identify the equilibrium-climate-sensitivity range.
\end{exercise}

\begin{exercise}
Derive the cost-effectiveness ratio for adaptation versus mitigation investment in a given region, given the climate-impact projection and the mitigation-cost projection.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the global annual investment required to achieve net-zero emissions by 2050 and compare to current global energy and infrastructure investment.
\end{exercise}

\begin{exercise}
Estimate the global land area required for the maximum credible bioenergy-with-carbon-capture-and-storage (BECCS) deployment and identify the conflict with agriculture, biodiversity, and other land uses.
\end{exercise}

\begin{exercise}
Estimate the climate-adaptation finance gap for developing countries and identify the engineering and institutional implications of closing it.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate a global net-zero trajectory under different technology-deployment, finance, and policy assumptions. Identify the trajectories that meet the $1.5\degC$ target versus those that exceed it.
\end{exercise}

\begin{exercise}
Simulate the climate-system response to a stratospheric aerosol injection deployment of varying intensity and duration. Identify the system-response timescales and the consequences of unexpected termination.
\end{exercise}

\begin{exercise}
Simulate the regional climate-adaptation requirements under different warming scenarios for a specific region (e.g., a coastal city; a major agricultural region; a mountain water-supply region).
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a national net-zero strategy for a specified country. Specify the sectoral decarbonisation pathways, the technology investments, the policy instruments, the financing, the just-transition support.
\end{exercise}

\begin{exercise}
Design a regional climate-adaptation programme for a specified vulnerable region. Specify the engineering interventions across sectors, the financing structure, the institutional coordination, the monitoring-and-evaluation.
\end{exercise}

\begin{exercise}
Design a governance framework for solar-radiation-management research that supports credible technical assessment without enabling premature deployment. Specify the research scope, the international coordination, the decision-rules for any deployment consideration, the safeguards against unilateral action.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: solar-radiation management research is the appropriate engineering response to the credible probability that mitigation alone will not contain warming to safe levels. Argue the opposite. Identify the engineering, ethical, and governance considerations.
\end{exercise}

\begin{exercise}
Discuss the Montreal Protocol as a model for climate-engineering governance. Identify the structural features that supported success; identify the differences with the climate-change case that limit the analogy.
\end{exercise}

\begin{exercise}
Argue: the engineering profession's responsibility includes the engagement with the political-economic context that determines whether climate-engineering plans become reality; the technical analysis is necessary but not sufficient. Argue the opposite. Identify the implications for engineering-professional practice.
\end{exercise}

