% Volume 12: Civilization
% Opener: tagline + scope. Sits immediately after the \part{} declaration.

\vspace*{1.5cm}
\begin{center}
\itshape\large
Discipline becomes civilization
\end{center}
\vspace{0.5cm}

\noindent
This volume is what the previous eleven were for. The mathematics of Volume I, the materials of Volumes II and V, the forces of Volume III, the energy of Volume IV, the biology of Volume VI, the information of Volume VII, the machines of Volume VIII, the systems of Volume IX, the failure literacy of Volume X, the design discipline of Volume XI: all of these cash out at the scale of civilization. The artifact is a part of an infrastructure; the infrastructure is a part of a system-of-systems; the system-of-systems is what humans live inside. Volume XII is the engineering treatment of what humans live inside.

The volume's central claim is that engineering at civilization scale is engineering, not policy. The civil engineer who builds a bridge has applied technical discipline to a specific structure; the civilizational engineer who decides where bridges go, who pays, who maintains, who is displaced, and what the bridges connect to is applying the same discipline at the scale where engineering meets political economy. The two are continuous; the chapter's central insistence is that engineers who think their work ends at the bridge are mistaken about what their work is.

The arc of the chapters: Chapter 1 frames engineering at civilization scale and what changes when the scale changes. Chapters 2 through 8 cover the seven canonical civilization-scale engineered systems: energy, water, transport, food, healthcare, buildings and cities, communications and digital infrastructure. Chapter 9 covers defence and security as engineered systems, included because the project's wide stance does not let it be omitted; defence procurement and command-and-control are civilization-scale engineering and the dual-use considerations they raise are part of the engineering profession's responsibility. Chapter 10 covers space infrastructure: orbital mechanics, launch capacity, satellites and constellations, and the engineering of the part of civilization that now sits above the atmosphere. Chapter 11 covers climate engineering, mitigation, and adaptation, the contemporary frontier of civilization-scale engineering work. Chapter 12 covers planetary boundaries and biosphere engineering, where engineering meets the limits of the planet. Chapter 13 covers the political economy of engineered life: the engineer as one of many actors, with the discipline to do the technical work and the responsibility to engage with the political-economic context that determines whether the work serves civilisation or harms it. Chapter 14 closes the entire book: the future of the artificial.

The reader who finishes this volume has finished the book. The reader has the technical literacy of an engineer, the failure literacy of an honest one, the design discipline of a capable one, and the civilizational understanding that distinguishes engineering as a profession with collective responsibility from engineering as a technical activity for hire. The volume is openly normative where the project's editorial decisions demanded; it is also openly modest about what engineers can decide alone. The reader is asked to engage with civilization-scale engineering as a substantive participant rather than as an observer.

The volume's archetypes are scaling, ethics, balance, and (in the chapter on planetary boundaries) failure. The cases draw from across the engineered systems that constitute modern civilisation; the discipline is the integration of engineering judgment with civilizational responsibility.

\vspace{1cm}
