\chapter{Equations of state, phase changes}
\label{vol04:ch02}

\epigraph{The pressure of a perfect gas at constant temperature varies
inversely as the volume which it occupies.}{The eighteenth-century
statement of Boyle's law, restated and extended by Sadi Carnot,
\emph{Reflections on the Motive Power of Fire} (1824), in the Mendoza
translation \cite[p.~14]{hist:carnot1824}.}

\chapmeta{Half-life: foundational. Archetypes: stability, scaling. Exercise target: 30. Project track: trace a real refrigerant cycle on a $p$-$h$ diagram. Hazard class: low (chart and tabular work). Reader path default: standard, with sections explicitly tagged. Named cases: none required in this chapter; cavitation, water-hammer, and BLEVE failure modes are described by mechanism.}

\noindent
An equation of state relates the thermodynamic properties of a
substance to one another. The first law of chapter~1 told us what is
conserved; the equation of state tells us, for any state we ask about,
what the temperature is given the pressure and the volume, or the
internal energy given the temperature, or the enthalpy along a
saturation curve. The chapter develops the ideal-gas equation, real-gas
extensions, the phase diagrams and saturation states that underwrite
process engineering, and the failure section closes with the working
catalogue of phase-transition accidents.

\section{Ideal gas}\pathtag{core}

The ideal-gas equation of state,
\[
pV = nRT,
\]
is the simplest and most useful equation of state in engineering
practice. The four state variables (pressure $p$, volume $V$, mole
number $n$, temperature $T$) are linked by a single relation, with
$R = 8.314\,\text{J}/(\text{mol}\cdot\si{\kelvin})$ the universal gas
constant. Per unit mass, the equation reads $pv = R_s T$ with $R_s =
R/M$ and $M$ the molar mass: for air, $R_s = 287\,\text{J}/(\text{kg}
\cdot\si{\kelvin})$; for steam at temperatures well above the critical
point, $R_s = 462\,\text{J}/(\text{kg}\cdot\si{\kelvin})$.

The ideal-gas law derives from two microscopic assumptions: the gas
molecules occupy negligible volume relative to the container, and the
intermolecular forces are negligible except during brief collisions.
The two assumptions hold to within a few percent for diatomic and
small molecules (air, nitrogen, oxygen, hydrogen) at pressures up to a
few atmospheres and temperatures well above their boiling points. They
fail for dense gases approaching critical conditions, for polar gases
(water vapour near saturation, ammonia), for large molecules
(hydrocarbons heavier than methane near liquefaction), and for any
gas near a phase transition.

The internal energy of an ideal gas depends only on temperature:
$dU = n c_v(T) \,dT$. The enthalpy likewise depends only on
temperature: $dH = n c_p(T) \,dT$. For monatomic ideal gases (argon,
helium, neon), $c_v = (3/2) R$ and $c_p = (5/2) R$ from equipartition,
giving $\gamma = c_p/c_v = 5/3 \approx 1.67$. For diatomic ideal gases
at temperatures where rotational modes are active but vibrational
modes are frozen out (air, nitrogen, oxygen near room temperature),
$c_v = (5/2) R$ and $c_p = (7/2) R$, giving $\gamma = 7/5 = 1.4$. For
polyatomic gases with rotational and partial vibrational excitation
(steam, carbon dioxide at room temperature), $\gamma$ falls toward
$\sim 1.3$. At high temperatures where additional internal modes
activate, $\gamma$ falls further; at very low temperatures where
rotational modes freeze, $\gamma$ rises toward $5/3$. The
temperature-dependence of $\gamma$ is the working signal that an
ideal-gas approximation is breaking down for reasons of internal
degrees of freedom rather than non-ideality.

A working engineer needs a quick test for when the ideal-gas law is
safe, and the reduced coordinates introduced with the compressibility
chart below give one: ideality is reliable to better than a percent when
the reduced pressure is low ($p_r < 0.1$) or the reduced temperature is
high ($T_r > 2$), and it grows untrustworthy as the state approaches the
saturation curve or the critical point. For air at room conditions the
reduced pressure is $p_r = 0.101/3.77 = 0.027$ and the reduced
temperature is $T_r = 293/132.5 = 2.21$, both deep in the safe band,
which is why the ideal-gas law serves so well for atmospheric air. Steam
at a turbine inlet, by contrast, sits at $T_r \approx 0.7$ to $1.0$ and
is a polar vapour near its saturation curve, exactly where the law fails;
the steam tables exist because the ideal-gas law cannot carry that
calculation.

Two numerical reflexes follow from the equation and pay for themselves.
The molar volume of any ideal gas at $0\,\degC$ and one atmosphere is
$v_m = RT/p = (8.314)(273.15)/101{,}325 = 22.4\,\text{L}/\text{mol}$, a
constant worth memorising; at $25\,\degC$ it is $24.5\,\text{L}/\text{mol}$.
And for a fixed mass in a closed vessel the law collapses to the
combined gas law $p_1 V_1/T_1 = p_2 V_2/T_2$ in absolute temperature, the
form that governs everything from a warmed sealed bottle to the pressure
swing of a gas cylinder left in the sun. The ideal-gas approximation is
the engineering default for gases at ambient and modest conditions. Its
limitations are catalogued in the next section.

\section{Real-gas equations}\pathtag{standard}

A real gas departs from the ideal-gas equation for two reasons: the
finite size of the molecules makes the available volume less than the
container volume, and the attractive intermolecular forces reduce the
pressure exerted on the wall. The simplest correction is the
\engterm{van der Waals equation},
\[
\left(p + \frac{a n^2}{V^2}\right)(V - nb) = nRT,
\]
with $a$ accounting for attractive forces and $b$ for the excluded
volume of the molecules. For nitrogen, $a = 0.137\,\text{Pa}\cdot
\text{m}^6/\text{mol}^2$ and $b = 3.87 \times 10^{-5}\,\text{m}^3/
\text{mol}$. The van der Waals equation predicts a critical point
(the temperature above which the gas cannot be liquefied by pressure
alone) and a coexistence region between gas and liquid; it is the
simplest equation that exhibits real-gas phenomenology, although it
is quantitatively inaccurate near the critical point.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-vdw-loops.tex}

In reduced variables the van der Waals equation loses its substance
constants entirely and reads $p_r = 8T_r/(3v_r - 1) - 3/v_r^2$, the form
plotted in figure~\ref{fig:vol04ch02:vdw-loops}. Below the critical
temperature the isotherm develops a loop with a region of $(\partial
p/\partial v)_T > 0$, which is mechanically unstable: a fluid in that
branch would respond to a small compression by lowering its own pressure,
so any fluctuation runs away and the homogeneous state cannot persist. The
physical isotherm replaces the loop with a horizontal coexistence line at
the saturation pressure $p_{\text{sat}}$, joining a saturated liquid at
$v_f$ to a saturated vapour at $v_g$.

\engterm{The Maxwell equal-area construction} fixes the height of that line,
and it is worth deriving rather than asserting, because the construction is
the same equal-Gibbs-energy condition that drives the Clausius-Clapeyron
relation, seen on the $p$-$v$ plane. Equilibrium between the two phases
requires equal specific Gibbs free energy, $g_f = g_g$, at the common
pressure and temperature. Along the isotherm $dg = -s\,dT + v\,dp = v\,dp$
at constant $T$, so integrating $g$ from the liquid endpoint to the vapour
endpoint along the van der Waals loop and demanding that the result return
to the same value gives
\[
g_g - g_f = \int_{f}^{g} v\,dp = 0
\quad\text{along the isotherm.}
\]
Integrating by parts, $\int_f^g v\,dp = [p\,v]_f^g - \int_f^g p\,dv =
p_{\text{sat}}(v_g - v_f) - \int_{v_f}^{v_g} p(v)\,dv$, because pressure is
equal at the two saturated endpoints. Setting this to zero,
\[
\int_{v_f}^{v_g} p(v)\,dv = p_{\text{sat}}\,(v_g - v_f),
\]
which states that the area under the true van der Waals loop between $v_f$
and $v_g$ equals the area of the rectangle under the horizontal line at
$p_{\text{sat}}$. The difference between the loop and the line above the
line (the liquid-side lobe) must therefore equal the difference below it
(the vapour-side lobe): the coexistence line is drawn so that it cuts equal
areas from the two lobes, the statement
figure~\ref{fig:vol04ch02:maxwell-equal-area} makes visible. The
construction is solved by adjusting $p_{\text{sat}}$ until the two lobes
balance, which simultaneously locates $v_f$ and $v_g$ as the two outer
intersections of the line with the loop. The routine
\texttt{code/vdw\_isotherms.py} performs that construction numerically by
bisecting on $p_{\text{sat}}$; at $T_r = 0.90$ it returns a reduced
saturation pressure of $0.647$ with $v_{f,r} \approx 0.60$ and $v_{g,r}
\approx 2.35$, the values the figure marks.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-maxwell-equal-area.tex}

The width of the coexistence line, $v_g - v_f$, is the volume change of
vapourisation that enters the Clausius-Clapeyron relation later in the
chapter, and it collapses to zero at the critical point where the loop
flattens into the horizontal inflexion. The construction therefore ties the
three pictures together: the $p$-$v$ loop and its equal-area line, the
saturation curve on the $p$-$T$ plane, and the latent heat that the
Clausius-Clapeyron slope carries. The van der Waals numbers are
quantitatively poor near the critical point, but the construction itself is
exact for any equation of state with a sub-critical loop, and a fitted cubic
run through the same bisection returns saturation pressures good to a few
percent.

More accurate real-gas equations used in process engineering include:

\begin{itemize}[leftmargin=*]
  \item \engterm{Redlich-Kwong} (1949): the working real-gas equation
    for non-polar gases at moderate pressures.
  \item \engterm{Soave-Redlich-Kwong} (SRK, 1972) and
    \engterm{Peng-Robinson} (PR, 1976): modifications used universally
    in the petroleum and natural-gas industries; both fit volumetric
    and vapour-liquid equilibrium data to within a few percent for
    hydrocarbon mixtures.
  \item \engterm{Virial equation}: a power-series expansion in
    density, $pv/RT = 1 + B(T)/v + C(T)/v^2 + \ldots$, with
    coefficients tabulated from experimental data. The truncated
    virial equation is accurate at low to moderate densities and is
    used in fundamental thermophysical work.
  \item \engterm{Benedict-Webb-Rubin} and modern multiparameter
    equations: reference-quality fits used in tabulated property data
    (NIST REFPROP, Wagner-Pruss water EoS) with accuracies at the
    $0.1\,\%$ level over wide ranges \cite{text:smith-van-ness-thermo}.
\end{itemize}

For process design, the working approach is to consult an equation of
state appropriate to the substance and conditions, either through a
property-tables look-up or through a computer-aided thermodynamics
package (Aspen Plus, ChemCAD, MATLAB Thermo libraries, NIST REFPROP).

\engterm{The compressibility factor}. A common way of expressing
real-gas behaviour is the compressibility factor,
\[
Z = \frac{pv}{R_s T},
\]
which equals $1$ for an ideal gas and departs from $1$ for a real
one. The \engterm{generalised compressibility chart} plots $Z$
against reduced pressure $p_r = p/p_c$ and reduced temperature $T_r =
T/T_c$ (with $p_c$, $T_c$ the critical pressure and temperature of
the substance), giving a universal curve that applies to most
non-polar gases. The chart is the engineer's quick check: at $p_r <
0.1$ or $T_r > 2$, $Z \approx 1$ and the ideal-gas approximation is
safe; at $0.5 < T_r < 1.5$ and $p_r$ of order unity, $Z$ varies from
$\sim 0.3$ (compressed gas at low temperature) to $\sim 1.5$ (gas
near critical conditions); figure~\ref{fig:vol04ch02:compressibility-chart}
plots the family of isotherms. The principle of corresponding states
underwriting the chart is itself an experimental observation, not a
derivable theorem, and is exact only for substances obeying the same
intermolecular potential.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-compressibility-chart.tex}

The departure of $Z$ from unity is quick to size with a real-gas
equation. The routine \texttt{code/ideal\_vs\_real\_gas.py} evaluates
both the ideal-gas pressure and the van der Waals pressure for a given
temperature and molar volume and reports the compressibility factor and
the percent error of the ideal-gas assumption; for nitrogen at
$300\,\si{\kelvin}$ and a molar volume of $5\times 10^{-4}\,\text{m}^3/
\text{mol}$ it returns $Z = 0.97$ (a $2.7\,\%$ error), while for carbon
dioxide near its critical region it returns $Z = 0.38$ (a $165\,\%$
error), the swing the chart makes visible. The critical constants used
to form $p_r$ and $T_r$ for the common engineering substances are
tabulated in \texttt{data/critical\_constants.csv}.

\section{The law of corresponding states}\pathtag{standard}

The compressibility chart works because of a structural fact about the
cubic equations of state, and the fact is worth deriving rather than
asserting. We take the van der Waals equation and rewrite it in the
reduced variables $p_r = p/p_c$, $T_r = T/T_c$, and $v_r = v/v_c$, using
the critical-point relations derived in the exercises, $v_c = 3b$, $T_c =
8a/(27Rb)$, and $p_c = a/(27b^2)$. Substituting $p = p_r\,p_c$, $T =
T_r\,T_c$, and $v = v_r\,v_c$ into the molar form $p = RT/(v-b) - a/v^2$
gives
\[
p_r\,\frac{a}{27b^2} = \frac{R\,T_r\,(8a/27Rb)}{3b\,v_r - b} -
\frac{a}{9b^2\,v_r^2}.
\]
Every occurrence of $a$ and $b$ cancels once the common factor $a/(27b^2)$
is divided through, and the equation collapses to
\[
p_r = \frac{8\,T_r}{3v_r - 1} - \frac{3}{v_r^2},
\]
the reduced form already plotted in figure~\ref{fig:vol04ch02:vdw-loops}.
The substance constants have vanished. Two different gases at the same
reduced temperature and reduced pressure occupy the same reduced volume
and therefore the same compressibility factor, because the reduced
equation that relates the three reduced variables carries no memory of
which gas it describes. This is the \engterm{law of corresponding states}:
the equation of state, written in reduced coordinates, is universal.

The law is exact for any fluid whose intermolecular potential has the
same shape and differs only in its energy and length scales, because the
two scale parameters are exactly what the reduction divides out. Real
potentials are not all the same shape, so the law is an approximation, but
a good one for the small non-polar molecules whose potentials are close to
a common Lennard-Jones form. It is the reason a single generalised
compressibility chart, indexed by $(p_r, T_r)$, serves for nitrogen,
methane, argon, and most light hydrocarbons at once, and the reason the
chart fails for water and ammonia, whose hydrogen bonding gives them a
potential of a different shape. The acentric factor of the mastery box is
the first correction to the law: a third reduced parameter that captures
the leading departure of a real potential from the simple-fluid shape, so
that a two-parameter chart indexed by $(p_r, T_r, \omega)$ recovers most
fluids to a few percent. The van der Waals equation gets the structure of
the law exactly right and the numbers wrong, predicting the same critical
compressibility $Z_c = 3/8$ for every substance against a measured cluster
near $0.27$; the structure is what survives into the industrial cubics,
the numbers are what they refit.

\section{Thermodynamic property relations}\pathtag{mastery}

A property table hands the engineer enthalpy and entropy directly, but the
engineer who works from an equation of state must build them, and the
bridge is a set of identities that turn the awkward derivatives of energy
and entropy into the measurable derivatives of $p$, $v$, and $T$. We
develop the identities here because the departure functions and the
fugacity of the next sections rest on them, and because a self-learner who
has only seen the equations quoted gains by watching them fall out of the
first and second laws.

The combined first and second laws for a closed simple compressible system,
written per mole, are $du = T\,ds - p\,dv$. Adding the differentials of the
three derived potentials gives the four \engterm{fundamental relations},
\begin{align*}
du &= T\,ds - p\,dv, \\
dh &= T\,ds + v\,dp, \\
da &= -s\,dT - p\,dv, \\
dg &= -s\,dT + v\,dp,
\end{align*}
where $h = u + pv$ is the enthalpy, $a = u - Ts$ the Helmholtz free
energy, and $g = h - Ts$ the Gibbs free energy. Each relation expresses an
exact differential of a state function, so the mixed second partial
derivatives are equal, and that equality applied to the four relations
gives the four \engterm{Maxwell relations},
\[
\left(\frac{\partial T}{\partial v}\right)_s = -\left(\frac{\partial
p}{\partial s}\right)_v, \quad
\left(\frac{\partial T}{\partial p}\right)_s = \left(\frac{\partial
v}{\partial s}\right)_p, \quad
\left(\frac{\partial s}{\partial v}\right)_T = \left(\frac{\partial
p}{\partial T}\right)_v, \quad
\left(\frac{\partial s}{\partial p}\right)_T = -\left(\frac{\partial
v}{\partial T}\right)_p.
\]
The third of these is the one that earns its keep in this chapter. Its left
side, the change of entropy with volume at fixed temperature, is not
something a measurement reaches directly; its right side, the change of
pressure with temperature at fixed volume, is exactly what an equation of
state delivers. The relation lets us replace an entropy derivative by an
equation-of-state derivative wherever the two appear together, and that
substitution is the move that makes the departure functions computable.

Two consequences follow at once. Writing $s$ as a function of $T$ and $v$,
$ds = (\partial s/\partial T)_v\,dT + (\partial s/\partial v)_T\,dv =
(c_v/T)\,dT + (\partial p/\partial T)_v\,dv$, where the first coefficient
uses the definition of $c_v$ and the second uses the third Maxwell
relation. Substituting this into $du = T\,ds - p\,dv$ gives the energy
equation
\[
\left(\frac{\partial u}{\partial v}\right)_T = T\left(\frac{\partial
p}{\partial T}\right)_v - p,
\]
the quantity the enthalpy-departure derivation called the bracket. For an
ideal gas $p = RT/v$ makes the right side vanish, which is the formal
statement that the internal energy of an ideal gas depends on temperature
alone. For a real gas the bracket measures how far the internal energy
strays from that ideal, and it is positive wherever attraction dominates,
because pulling the molecules apart against their mutual attraction stores
energy. This single identity, with the third Maxwell relation behind it, is
the engine of the departure functions.

\section{Departure functions: building real properties from an equation of
state}\pathtag{standard}

A \engterm{departure function}, also called a residual property, is the
difference between a real fluid's property and the ideal-gas property at the
same temperature and pressure. The strategy of property calculation from an
equation of state is to compute the easy ideal-gas value from the
temperature alone and then add the departure, which the equation of state
supplies. The chapter has already derived the enthalpy departure for the
van der Waals fluid in a worked example; here we develop the general result
and the entropy departure that goes with it, because a cycle calculation
needs both.

The general enthalpy departure follows from integrating the energy equation
along an isotherm from the ideal-gas limit ($v\to\infty$) to the real molar
volume. From $h = u + pv$ and the energy equation,
\[
h - h^{\text{ideal}} = \int_{\infty}^{v}\left[T\left(\frac{\partial
p}{\partial T}\right)_v - p\right]dv + (pv - RT),
\]
the $pv - RT$ term supplying the difference in the $pv$ products between the
real state and the ideal-gas state at the same temperature. The entropy
departure follows the same path from $ds = (c_v/T)\,dT + (\partial
p/\partial T)_v\,dv$. Comparing the real fluid to an ideal gas brought to
the same final temperature and pressure, the temperature integral of
$c_v/T$ cancels between the two because both share the same ideal-gas heat
capacity, and what remains is the volume integral of the difference in
$(\partial p/\partial T)_v$ between the real fluid and the ideal gas,
\[
s - s^{\text{ideal}} = \int_{\infty}^{v}\left[\left(\frac{\partial
p}{\partial T}\right)_v - \frac{R}{v}\right]dv + R\ln\frac{pv}{RT},
\]
the final logarithm correcting for the fact that the ideal-gas reference is
taken at the same pressure, not the same volume, so its molar volume is
$RT/p$ rather than $v$.

We carry both integrals through for the van der Waals fluid, for which
$(\partial p/\partial T)_v = R/(v-b)$. The enthalpy departure reproduces
the worked-example result,
\[
h - h^{\text{ideal}} = \frac{RTb}{v-b} - \frac{2a}{v},
\]
and the entropy departure is
\[
s - s^{\text{ideal}} = \int_{\infty}^{v}\left[\frac{R}{v-b} -
\frac{R}{v}\right]dv + R\ln\frac{pv}{RT} = R\ln\frac{v-b}{v} +
R\ln\frac{pv}{RT} = R\ln\frac{p(v-b)}{RT}.
\]
The first integral is elementary because $R/(v-b) - R/v$ integrates to
$R\ln[(v-b)/v]$ with the value zero at the upper limit $v\to\infty$, and
the algebra collapses the two logarithms. The entropy departure is
negative wherever $p(v-b) < RT$, which is the rule for a dense gas: the
real fluid has less entropy than an ideal gas at the same conditions
because its molecules, confined by their finite size and held by their
attraction, sample fewer configurations. The two departures together let
the engineer assemble a real enthalpy and entropy from an ideal-gas
calculation, $h = h^{\text{ideal}}(T) + (h - h^{\text{ideal}})$ and
likewise for $s$, which is exactly what a process simulator does at every
flowsheet node before it can apply an energy balance. The generalised
charts of figure~\ref{fig:vol04ch02:departure-functions} present the same
departures in corresponding-states form, indexed by $(p_r, T_r)$, for an
engineer working from reduced coordinates rather than from a specific
equation of state.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-departure-functions.tex}

\section{Fugacity and the escaping tendency of a real gas}\pathtag{standard}

Phase and chemical equilibrium are governed by the Gibbs free energy, and
the chapter's Maxwell construction has already used the condition $g_f =
g_g$ to locate the saturation line. The working tool for carrying that
condition to a real gas is the \engterm{fugacity}, the property that plays
for a real gas the role the pressure plays for an ideal one. We define it
so that the chemical potential keeps its ideal-gas form.

For an ideal gas at constant temperature, $dg = v\,dp = (RT/p)\,dp = RT\,d
\ln p$, so the molar Gibbs energy is $g = g^{\circ}(T) + RT\ln(p/p^{\circ})$
with $p^{\circ}$ a reference pressure. A real gas does not obey this,
because $v \neq RT/p$. The fugacity $f$ is defined to preserve the form,
\[
g = g^{\circ}(T) + RT\ln\frac{f}{p^{\circ}}, \qquad \lim_{p\to 0}\frac{f}{p}
= 1,
\]
the limit fixing the fugacity to equal the pressure as the gas thins toward
ideality. The fugacity is the corrected, thermodynamically consistent
pressure: the quantity an engineer substitutes for pressure in any
equilibrium relation to make it hold for a real gas. The ratio $\phi = f/p$
is the \engterm{fugacity coefficient}, equal to one for an ideal gas and
departing from one in measure of the gas's non-ideality, plotted against
reduced pressure in figure~\ref{fig:vol04ch02:fugacity-coefficient}.

The fugacity coefficient is computable from the equation of state. At
constant temperature the definition gives $RT\,d\ln f = dg = v\,dp$, while
for an ideal gas $RT\,d\ln p = (RT/p)\,dp$. Subtracting,
\[
RT\,d\ln\frac{f}{p} = \left(v - \frac{RT}{p}\right)dp,
\]
and integrating from zero pressure, where $\phi = 1$, up to the pressure of
interest,
\[
\ln\phi = \ln\frac{f}{p} = \frac{1}{RT}\int_{0}^{p}\left(v -
\frac{RT}{p}\right)dp = \int_{0}^{p}\frac{Z - 1}{p}\,dp,
\]
where the last form, using $Zv = RT/p$ to write $v - RT/p = (Z-1)RT/p$,
expresses the fugacity coefficient as an integral of the compressibility
factor's departure from unity over the pressure path. The integral is
positive where $Z > 1$ (repulsion dominant, the gas harder to compress than
ideal, $\phi > 1$) and negative where $Z < 1$ (attraction dominant, the gas
easier to compress, $\phi < 1$), which is why the fugacity-coefficient chart
dips below one near the critical temperature and climbs above one at high
reduced pressure and temperature, mirroring the compressibility chart that
feeds it. The fugacity is the quantity equated across a phase boundary,
$f_f = f_g$ at saturation, the real-gas refinement of the equal-Gibbs-energy
condition, and it is the quantity a vapour-liquid equilibrium calculation in
a process simulator equates component by component to split a mixture
between its phases.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-fugacity-coefficient.tex}

\section{Liquids and solids: incompressibility approximation}\pathtag{standard}

Liquids and solids are nearly incompressible at engineering
pressures: doubling the pressure on liquid water at room temperature
changes its volume by less than $0.05\,\%$. The
\engterm{incompressibility approximation} treats the specific volume
as a constant equal to its value at a reference state, often the
saturated-liquid value at the operating temperature. With $v$
constant, several relations simplify:

\begin{itemize}[leftmargin=*]
  \item Internal energy depends only on temperature: $du = c \,dT$.
  \item Enthalpy varies with both temperature and pressure, but the
    pressure dependence is small: $dh = c\,dT + v\,dp$. For most
    engineering work, the pressure term is negligible compared to the
    temperature term except at very high pressures (hydraulic systems
    above $50\,\text{MPa}$, deep-ocean engineering).
  \item Entropy depends only on temperature: $ds = c\,dT/T$, with $c$
    treated as the single specific heat ($c_p \approx c_v$ for
    incompressible substances).
\end{itemize}

For working calculations, liquid water at moderate temperatures has
$c \approx 4.18\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$ and $v
\approx 1.0 \times 10^{-3}\,\text{m}^3/\text{kg}$. Other engineering
liquids (oils, refrigerant condensate, alcohols) have $c$ in the
range $1.5$ to $2.5\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$ and $v$
in the range $1.0$ to $1.5 \times 10^{-3}\,\text{m}^3/\text{kg}$.
Solids have $c$ from $\sim 0.4\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$
for heavy metals (steel, lead) to $\sim 1.0\,\text{kJ}/(\text{kg}
\cdot\si{\kelvin})$ for light materials (aluminium, concrete) and
$\sim 2.0\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$ for wood and
polymers.

The size of the approximation is worth a number. The fractional volume
change of a liquid under pressure is set by its bulk modulus $K$ through
$\Delta v/v = -\Delta p/K$. Liquid water has $K \approx 2.2\,\text{GPa}$,
so raising the pressure by $10\,\text{MPa}$ (a hundred atmospheres,
typical of a high-pressure pump discharge) compresses the water by
$\Delta v/v = -10/2200 = -0.45\,\%$, a change a design can usually
ignore. The same bulk modulus, divided by the density, sets the speed of
sound, $c = \sqrt{K/\rho} = \sqrt{2.2\times 10^9/1000} = 1480\,\text{m}/
\text{s}$, the number that reappears in the water-hammer wave speed of
the failure section. The incompressibility approximation discards the
volume change but the same bulk modulus that makes the change small makes
the sound speed large, which is why a liquid that is nearly
incompressible for steady calculations transmits a pressure transient
almost instantly.

The incompressibility approximation breaks down near critical
conditions, in deep-pressure engineering (hydrostatic pressures
exceeding several MPa where small compressibility produces
significant energy storage), and in high-strain-rate phenomena
(shock waves, water-hammer transients) where the speed of sound and
the bulk modulus of the liquid matter. The water-hammer case is the
cleanest illustration: the steady-flow analysis treats the water as
incompressible and conserves volume, while the transient analysis
restores the compressibility, because the entire pressure pulse is the
liquid's small elastic give propagating at the sound speed. The two
treatments are not in conflict; they answer different questions about the
same fluid, and the engineer chooses by the time scale of the problem.

The one place the incompressibility of a liquid becomes a hazard rather than
a convenience is a sealed vessel filled liquid-full and then warmed. A liquid
expands with temperature at a volume thermal-expansion coefficient $\beta =
(1/v)(\partial v/\partial T)_p$, about $2.1\times 10^{-4}\,\si{\kelvin}^{-1}$
for water near room temperature. If the liquid has a vapour space to expand
into, the expansion is harmless; if the vessel is liquid-full and rigid, the
liquid cannot expand, and the frustrated expansion is taken up as a pressure
rise against the bulk modulus. Holding the volume fixed and combining the
thermal expansion with the compressibility gives the pressure rise per degree
directly,
\[
\left(\frac{\partial p}{\partial T}\right)_v = \frac{\beta}{\kappa} = \beta
K,
\]
where $\kappa = 1/K$ is the isothermal compressibility. For water this is
$(\partial p/\partial T)_v = (2.1\times 10^{-4})(2.2\times 10^9) = 4.6\times
10^5\,\text{Pa}/\si{\kelvin}$, that is $0.46\,\text{MPa}$ for every degree of
warming. A liquid-full sealed pipe or vessel warmed by ten degrees would gain
$4.6\,\text{MPa}$, enough to burst most piping, which is why thermal-relief
valves are fitted to any line that can be valved off full of liquid and then
heated by the sun or by a hot process nearby, and why the propane case
study's $80\,\%$ fill limit exists: the vapour ullage is the compressible
volume that keeps the tank pressure on the gentle saturation curve rather
than on the steep liquid bulk modulus. The same relation read the other way
is a caution about the incompressibility approximation itself: treating a
liquid as strictly incompressible ($K\to\infty$) predicts an infinite
pressure rise on constrained heating, so the small compressibility the
approximation discards is exactly the quantity that sets this pressure, one
more case where the term dropped for steady calculations governs the
transient the vessel actually fails by.

\section{Phase diagrams: $p$-$V$-$T$, $T$-$s$, $h$-$s$}\pathtag{core}

A phase diagram displays the regions in property space where a
substance exists as solid, liquid, vapour, and supercritical fluid,
and the curves along which two or three phases coexist. The
engineering working diagrams plot pressure against specific volume
($p$-$v$), temperature against entropy ($T$-$s$), or enthalpy against
entropy ($h$-$s$, also called the Mollier diagram).

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-pvt-surface.tex}

Every equilibrium state of a simple compressible substance is a single
point on one three-dimensional $p$-$v$-$T$ surface
(figure~\ref{fig:vol04ch02:pvt-surface}). The working diagrams are
projections of that surface onto a pair of axes. The $p$-$T$ phase
diagram (figure~\ref{fig:vol04ch02:pt-phase-diagram}) is the projection
along the volume axis: the three coexistence curves (sublimation, fusion,
vapourisation) appear as lines, the triple point as a single point where
all three phases coexist, and the vapourisation curve ends at the
critical point.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-pt-phase-diagram.tex}

\engterm{The $p$-$v$ diagram} shows the substance's specific volume
as a function of pressure at constant temperature. The isotherms
above the critical temperature are monotonic in $v$. The isotherms
below the critical temperature pass through the two-phase region
where the substance exists as a saturated liquid plus saturated
vapour, with $v$ varying between the saturated-liquid value $v_f$
and the saturated-vapour value $v_g$
(figure~\ref{fig:vol04ch02:pv-isotherms-dome}). Inside the two-phase region,
the isotherm is horizontal (pressure equals the saturation pressure
for that temperature). The locus of critical conditions defines the
\engterm{critical point}: for water, $T_c = 647\,\si{\kelvin}$ and
$p_c = 22.06\,\text{MPa}$; for ammonia, $T_c = 405\,\si{\kelvin}$ and
$p_c = 11.3\,\text{MPa}$; for carbon dioxide, $T_c = 304\,\si{\kelvin}$
and $p_c = 7.4\,\text{MPa}$.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-pv-isotherms-dome.tex}

\engterm{The $T$-$s$ diagram} shows temperature against entropy. The
two-phase region is the area under a bell-shaped curve (the
\engterm{vapour dome}) whose peak is the critical point. Saturated
liquid lies along the left limb of the dome; saturated vapour along
the right limb. Isobars are horizontal lines inside the dome (constant
$T$ at saturation for given $p$) and curve upward outside. The $T$-$s$
diagram is the natural representation for Carnot, Rankine, and
refrigeration cycles because the area enclosed by a cycle on the
$T$-$s$ diagram equals the net heat absorbed in the cycle, and the
direction of the cycle around the diagram (clockwise for engines,
counter-clockwise for refrigeration) corresponds to the sign of net
work; figure~\ref{fig:vol04ch02:ts-dome-cycle} inscribes a steam-power
cycle on the dome.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-ts-dome-cycle.tex}

\engterm{The Mollier diagram ($h$-$s$)}. The Mollier diagram is the
working chart for steam-power and air-conditioning processes. The
$h$-$s$ representation makes process work and heat read directly off
the diagram: a turbine's work output is the vertical drop in $h$
between inlet and outlet, an isentropic expansion is a vertical line,
a heat-exchanger duty is the horizontal distance, and irreversibility
manifests as deviation from vertical or horizontal. The diagram's
practical utility is the reason Richard Mollier introduced it in 1904
for steam analysis; modern engineering still uses Mollier-style
charts in addition to (and sometimes in place of) digital property
look-ups.

The three diagrams represent the same thermodynamic surface in
different parameter pairs. A working engineer reads any of them with
the same speed; the choice between them is set by the process. For a
heat engine, $T$-$s$ makes the Carnot bound and the cycle's
irreversibility visible. For a flow process with shaft work, $h$-$s$
makes the work read directly. For a refrigerant compressor, the
$p$-$h$ diagram (pressure against enthalpy) is the canonical
representation, with the cycle drawn as a four-leg loop
(figure~\ref{fig:vol04ch02:ph-refrigerant-cycle}, developed with the
worked refrigeration example below).

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-ph-refrigerant-cycle.tex}

\begin{archetype}[Stability]
The two-phase region of a substance is the region of marginal
mechanical stability: $(\partial p/\partial v)_T = 0$ along the
saturation curve. A system poised in or near this region can
spontaneously redistribute between phases, releasing or absorbing
latent heat. The engineering of fluid systems is largely the
engineering of approach to and recovery from this stability boundary:
pumps avoid the boundary by maintaining suction-pressure margin,
boilers approach it deliberately to extract latent heat, and refrigerant
cycles cross it twice per loop. The stability archetype recurs
across the volume: the second law's $\dot{S}_{\text{gen}} \geq 0$
(Ch~1, Ch~4), the static-stability criteria for fluids (Vol~III), and
the dynamical-stability margins of feedback systems (Vol~IX).
\end{archetype}

\section{Saturation states and quality}\pathtag{core}

Inside the two-phase region, the substance exists as a mixture of
saturated liquid and saturated vapour at the same temperature and
pressure. The state is parameterised by the \engterm{quality} $x$,
defined as the mass fraction of vapour in the mixture:
\[
x = \frac{m_g}{m_g + m_f},
\]
where $m_g$ and $m_f$ are the masses of saturated vapour and
saturated liquid respectively. Quality $x = 0$ is saturated liquid;
$x = 1$ is saturated vapour; $0 < x < 1$ is a two-phase mixture.

All extensive properties of a two-phase mixture are weighted averages
of the saturated-liquid and saturated-vapour values:
\begin{align*}
v &= (1 - x) v_f + x v_g, \\
u &= (1 - x) u_f + x u_g, \\
h &= (1 - x) h_f + x h_g, \\
s &= (1 - x) s_f + x s_g.
\end{align*}
Each weighting is read graphically by the \engterm{lever rule}
(figure~\ref{fig:vol04ch02:quality-lever}): on a diagram with the
mixture property on the horizontal axis, the two-phase state divides the
tie line between $v_f$ and $v_g$ so that the vapour mass fraction equals
the segment from $v_f$ to the state divided by the full tie length.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-quality-lever.tex}
The saturated-liquid and saturated-vapour properties are tabulated as
functions of either temperature or pressure; the two are connected by
the \engterm{saturation curve} on the $p$-$T$ phase diagram. For
water, the saturation pressure at $100\,\degC$ is $101.325\,\text{kPa}$;
at $200\,\degC$, $1554\,\text{kPa}$; at $300\,\degC$, $8587\,\text{kPa}$.
The pressure rises rapidly with temperature, reflecting the latent
heat of vapourisation and the Clausius-Clapeyron relation
\[
\frac{dp_{\text{sat}}}{dT} = \frac{h_{fg}}{T(v_g - v_f)},
\]
which integrates (assuming $v_g \gg v_f$ and ideal-gas behaviour for
the vapour) to $\ln p \approx -h_{fg}/(R_s T) + C$, the form used in
saturation-pressure correlations. Plotting $\ln p_{\text{sat}}$ against
$1/T$ turns the relation into a near-straight line whose slope is
$-h_{fg}/R_s$ (figure~\ref{fig:vol04ch02:clausius-clapeyron}); the mild
curvature at the high-temperature end records the temperature dependence
of $h_{fg}$. The routine \texttt{code/clausius\_clapeyron.py} carries the
two-point correlation and the slope fit, predicting $p_{\text{sat}}$ at
$150\,\degC$ from the $100\,\degC$ anchor to within the steam table.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-clausius-clapeyron.tex}

The latent heat of vapourisation $h_{fg} = h_g - h_f$ depends
strongly on temperature, decreasing from $2501\,\text{kJ}/\text{kg}$
at $0\,\degC$ to $2257\,\text{kJ}/\text{kg}$ at $100\,\degC$ and
falling to zero at the critical point ($647\,\si{\kelvin}$). The
temperature dependence of $h_{fg}$ is responsible for the difference
between cycle calculations at different operating temperatures and
for the higher efficiency limits of high-temperature processes
(chapter~3 of this volume).

\engterm{Subcooled and superheated states}. Outside the two-phase
region, the substance is either subcooled liquid (temperature below
the saturation temperature at the given pressure) or superheated
vapour (temperature above the saturation temperature at the given
pressure). The subcooled-liquid properties are often approximated by
the saturated-liquid values at the same temperature (the
incompressibility approximation). The superheated-vapour properties
are tabulated in superheated-vapour tables or computed from a
real-gas equation of state.

The distance from the saturation curve is itself a working quantity, and it
carries two names an engineer must keep straight. The \engterm{degree of
subcooling} is the amount by which a liquid's temperature sits below the
saturation temperature at its pressure, $\Delta T_{\text{sub}} =
T_{\text{sat}}(p) - T$, and the \engterm{degree of superheat} is the amount by
which a vapour's temperature sits above it, $\Delta T_{\text{sup}} = T -
T_{\text{sat}}(p)$. The two are the safety margins the failure modes of the
chapter are written in. A pump's suction liquid must carry enough subcooling
that the pressure drop into the impeller eye does not take it to saturation and
cavitation, which is the net-positive-suction-head margin recast as a
temperature; a compressor's suction vapour must carry enough superheat that no
liquid droplets reach the valves and wash the oil, which is why a refrigeration
evaporator is run with a few degrees of exit superheat rather than to exactly
saturated vapour. The subcooling margin is quick to size in pressure. A liquid
at temperature $T$ has a saturation pressure $p_{\text{sat}}(T)$; the margin
before it flashes is the amount its actual pressure exceeds that saturation
pressure, $\Delta p = p - p_{\text{sat}}(T)$, and the Clausius-Clapeyron slope
converts a temperature subcooling into this pressure margin, $\Delta p \approx
(dp_{\text{sat}}/dT)\,\Delta T_{\text{sub}}$. For water at $90\,\degC$, where
the saturation-curve slope is about $4.6\,\text{kPa}/\si{\kelvin}$, a five-degree
subcooling buys only $23\,\text{kPa}$ of flash margin, which is why hot-water
pumps cavitate so readily and why the same five degrees at $20\,\degC$, where
the slope is a tenth as steep, buys almost nothing at all. The degree of
superheat plays the mirror role in a boiler: the superheat table's entries are
indexed by pressure and by the temperature above saturation, and the enthalpy a
superheater adds is the integral of $c_p$ across the superheat span, the third
leg of the boiler duty the worked boiler-profile example sizes. Reading a state
as subcooled or superheated, and by how much, is the first move in any table
look-up, because it decides which table the engineer opens: the compressed-liquid
table, the saturation table, or the superheated-vapour table.

\section{Properties tables and software}\pathtag{standard}

Engineering thermodynamics has, since the late nineteenth century,
relied on tabulated property data for working substances. The most
important tables for engineering practice are the \engterm{steam
tables}, which give the saturated and superheated properties of water
in the engineering ranges. Modern tables derive from the IAPWS-IF97
formulation (Industrial Formulation 1997) of the International
Association for the Properties of Water and Steam, which fits the
underlying thermodynamic surface to within experimental uncertainty
across the engineering range \cite{text:cengel-boles}.

Other working tables: refrigerant tables for R-134a, R-410A, R-1234yf,
and the older R-22 (now phased out); ammonia tables for industrial
refrigeration; carbon dioxide tables for transcritical heat-pump
cycles; air tables for high-temperature gas-turbine analysis; ideal-
gas tables for combustion products of common fuels.

The working alternative to tables is software. NIST REFPROP
implements reference-quality equations of state for over $150$ pure
fluids and many mixtures, with accuracies inherited from the
underlying experimental data. Aspen Plus, HYSYS, ChemCAD, and the
open-source CoolProp and Thermo libraries provide the same data with
process-flowsheet integration. For routine engineering, the choice
between table look-up and software is set by the process geometry: a
single point check uses a table, a process with iterative composition
or temperature solves uses software.

The reader must know how to interpolate in a table. Standard
interpolation is linear in the table's grid:
\[
y \approx y_1 + \frac{x - x_1}{x_2 - x_1}(y_2 - y_1),
\]
with the table's neighbouring values. A worked instance fixes the
mechanics. To find the enthalpy of saturated-liquid water at
$175\,\degC$ from the table rows at $150\,\degC$ ($h_f = 632.2$) and
$200\,\degC$ ($h_f = 852.3\,\text{kJ}/\text{kg}$), the fraction of the
way across the interval is $(175 - 150)/(200 - 150) = 0.50$, so
$h_f(175\,\degC) \approx 632.2 + 0.50(852.3 - 632.2) = 742.3\,\text{kJ}/
\text{kg}$. The table's own $175\,\degC$ entry is $741.0\,\text{kJ}/
\text{kg}$, so the linear interpolation overshoots by $0.2\,\%$, the
small convex curvature of $h_f$ against temperature. Two-dimensional
interpolation (in pressure and temperature, for example) requires four
corners and the bilinear formula: interpolate first in one variable along
each of the two bracketing rows, then interpolate between those two
results in the second variable. The interpolation introduces a small
error ($\sim 0.1\,\%$ for well-spaced tables in regions of smooth
behaviour, larger near the critical point or saturation curve, where the
properties bend sharply and a linear fit between widely spaced rows can
err by a percent or more), which is usually negligible compared to the
propagated uncertainties of the upstream measurements. The discipline is
to bracket the target with the nearest table rows, to interpolate rather
than round to the nearest entry, and to tighten the grid near the
saturation curve where the curvature is largest. The routine
\texttt{code/steam\_interpolate.py}
implements both the one-dimensional and the bilinear forms against a
saturated-water table (\texttt{data/water\_saturation.csv}); a companion
refrigerant table (\texttt{data/r134a\_saturation.csv}) drives the
refrigeration worked example below.

\begin{principle}[State is specified by two independent properties]
For a simple compressible substance with fixed composition, two
independent intensive properties specify the thermodynamic state. Any
other property follows from the equation of state and the tabulated
relations. Common working pairs: $(T, p)$ in single-phase regions,
$(T, x)$ or $(p, x)$ in the two-phase region, $(p, h)$ for flow
processes, $(p, s)$ for isentropic-end-state calculations. The choice
of pair depends on which two are measured or specified in the
process; the third and subsequent properties read off the table.
\end{principle}

\section{Worked examples: steam table use, refrigerant cycle}\pathtag{standard}

\subsection{Steam expanding through a turbine}

Steam enters a turbine at $5\,\text{MPa}$ and $400\,\degC$. The
exit is at $50\,\text{kPa}$. If the expansion is isentropic, what is
the exit state?

At the inlet, the steam table (superheated) gives $h_1 = 3196\,
\text{kJ}/\text{kg}$ and $s_1 = 6.65\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$. For an isentropic expansion to $50\,\text{kPa}$, the
exit entropy equals the inlet entropy: $s_2 = 6.65\,\text{kJ}/(\text{kg}
\cdot\si{\kelvin})$. At $50\,\text{kPa}$, the saturated-liquid and
saturated-vapour entropies are $s_f = 1.09\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$ and $s_g = 7.59\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$.
Since $s_f < s_2 < s_g$, the exit state is in the two-phase region.
The quality at exit is
\[
x_2 = \frac{s_2 - s_f}{s_g - s_f} = \frac{6.65 - 1.09}{7.59 - 1.09} =
0.855.
\]
The exit enthalpy is $h_2 = h_f + x_2 (h_g - h_f) = 340.5 + 0.855
\times 2305 = 2311\,\text{kJ}/\text{kg}$. The specific work output is
$w = h_1 - h_2 = 3196 - 2311 = 885\,\text{kJ}/\text{kg}$. At a mass
flow rate of $10\,\text{kg}/\text{s}$, the turbine power is $8.85\,
\text{MW}$.

The exit quality of $0.855$ is on the low end of acceptable for a
condensing steam turbine. Typical design limits the exit quality to
$\geq 0.9$ to limit blade erosion by liquid droplets. The remedy is
either reheating between two turbine stages or higher initial
superheat; chapter~3 of this volume develops the Rankine cycle's
working geometry.

\subsection{R-134a in a vapour-compression refrigeration cycle}

Refrigerant R-134a is compressed from a saturated-vapour state at
$-20\,\degC$ to $1.0\,\text{MPa}$, then condensed to a saturated-
liquid state at $1.0\,\text{MPa}$, then throttled to $-20\,\degC$,
then evaporated back to saturated vapour. The mass flow rate is $0.05
\,\text{kg}/\text{s}$. Assuming the compression is isentropic, find
the cooling capacity, the compressor power, and the coefficient of
performance.

State 1 (saturated vapour at $-20\,\degC$): from the R-134a tables,
$p_1 = 132.7\,\text{kPa}$, $h_1 = 235.3\,\text{kJ}/\text{kg}$, $s_1 =
0.9332\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$.

State 2 (after isentropic compression to $1.0\,\text{MPa}$): the
saturation temperature at $1.0\,\text{MPa}$ is $39.4\,\degC$; with
$s_2 = s_1 = 0.9332$, the superheated-vapour table at $1.0\,\text{MPa}$
gives $T_2 \approx 47\,\degC$, $h_2 \approx 277\,\text{kJ}/\text{kg}$.
The specific work of compression is $w_c = h_2 - h_1 = 277 - 235.3 =
41.7\,\text{kJ}/\text{kg}$.

State 3 (saturated liquid at $1.0\,\text{MPa}$): $h_3 = 107.3\,\text{kJ}/
\text{kg}$. The condenser duty (heat rejected) is $q_H = h_2 - h_3 =
277 - 107.3 = 169.7\,\text{kJ}/\text{kg}$.

State 4 (after throttling to $-20\,\degC$): the throttle is isenthalpic,
so $h_4 = h_3 = 107.3\,\text{kJ}/\text{kg}$. At $-20\,\degC$, $h_f =
24.3\,\text{kJ}/\text{kg}$, $h_g = 235.3\,\text{kJ}/\text{kg}$, so
$x_4 = (107.3 - 24.3)/(235.3 - 24.3) = 0.393$.

The evaporator duty (cooling capacity) is $q_L = h_1 - h_4 = 235.3 -
107.3 = 128.0\,\text{kJ}/\text{kg}$. The COP is
\[
\text{COP} = \frac{q_L}{w_c} = \frac{128.0}{41.7} = 3.07.
\]
At a mass flow of $0.05\,\text{kg}/\text{s}$, the cooling capacity is
$6.4\,\text{kW}$ and the compressor power is $2.1\,\text{kW}$.

The Carnot COP for the same temperature range ($T_c = 253\,\si{\kelvin}$,
$T_h = 313\,\si{\kelvin}$) is $T_c/(T_h - T_c) = 4.22$. The actual COP
is $73\,\%$ of the Carnot bound, reasonable for a single-stage cycle
with modest superheat and modest sub-cooling. The same arithmetic, taken
from the four state enthalpies, is collected in
\texttt{code/refrigeration\_cycle.py}, which returns the cooling
capacity, the work, the duty, and the fraction of the Carnot bound for
any measured cycle; the four state points of this example are the loop
plotted in figure~\ref{fig:vol04ch02:ph-refrigerant-cycle}.

\subsection{Van der Waals against the ideal gas for compressed carbon
dioxide}

The chart and the equations earn their keep when the substance is a
dense, near-critical gas. We size the error directly for carbon dioxide
at $310\,\si{\kelvin}$ confined to a molar volume of $v = 1.5\times
10^{-4}\,\text{m}^3/\text{mol}$, a state typical of a high-pressure
storage cylinder. The critical constants are $T_c = 304.1\,\si{\kelvin}$
and $p_c = 7.38\,\text{MPa}$ (\texttt{data/critical\_constants.csv}), so
the reduced temperature is $T_r = 310/304.1 = 1.02$, just above critical,
where the chart shows its deepest dip in $Z$.

The ideal-gas pressure is
\[
p_{\text{ideal}} = \frac{RT}{v} = \frac{(8.314)(310)}{1.5\times 10^{-4}}
= 1.718\times 10^7\,\text{Pa} = 17.2\,\text{MPa}.
\]
The van der Waals constants follow from the critical data,
$a = 27 R^2 T_c^2/(64 p_c) = 0.365\,\text{Pa}\cdot\text{m}^6/\text{mol}^2$
and $b = R T_c/(8 p_c) = 4.28\times 10^{-5}\,\text{m}^3/\text{mol}$. Then
\[
p_{\text{vdW}} = \frac{RT}{v - b} - \frac{a}{v^2}
= \frac{(8.314)(310)}{1.5\times 10^{-4} - 4.28\times 10^{-5}}
- \frac{0.365}{(1.5\times 10^{-4})^2}.
\]
The first term is $(2577.3)/(1.072\times 10^{-4}) = 2.404\times 10^7\,
\text{Pa}$; the second is $0.365/(2.25\times 10^{-8}) = 1.622\times 10^7\,
\text{Pa}$. The difference is $p_{\text{vdW}} = 7.82\times 10^6\,\text{Pa}
= 7.82\,\text{MPa}$. The compressibility factor is
\[
Z = \frac{p_{\text{vdW}}\, v}{RT} = \frac{(7.82\times 10^6)(1.5\times
10^{-4})}{(8.314)(310)} = \frac{1173}{2577} = 0.455.
\]
The ideal-gas law overstates the pressure by a factor of $17.2/7.82 =
2.2$, a $120\,\%$ error. The excluded-volume term ($V - nb$) raises the
pressure above ideal, the attraction term ($-a/v^2$) lowers it, and near
the critical point the attraction wins decisively. The lesson the chart
makes visual is here a single number: at $T_r \approx 1$ and a molar
volume a few times $b$, the ideal-gas pressure is not an approximation,
it is wrong by more than a factor of two, and only an equation of state
that carries both corrections recovers the real pressure. The routine
\texttt{code/ideal\_vs\_real\_gas.py} returns the same $Z$.

\subsection{Quality and the lever rule from a measured specific volume}

A rigid $0.10\,\text{m}^3$ vessel holds $20\,\text{kg}$ of water at
$200\,\degC$. We find the state and read the phase split off the lever
rule. The specific volume is fixed by the mass and the vessel,
\[
v = \frac{V}{m} = \frac{0.10}{20} = 5.0\times 10^{-3}\,\text{m}^3/
\text{kg}.
\]
From the saturated-water table at $200\,\degC$
(\texttt{data/water\_saturation.csv}), $v_f = 1.157\times 10^{-3}$ and
$v_g = 0.1272\,\text{m}^3/\text{kg}$. Since $v_f < v < v_g$, the water is
a two-phase mixture at the $200\,\degC$ saturation pressure of
$1555\,\text{kPa}$. The quality follows from inverting the mixture rule
$v = (1-x)v_f + x v_g$,
\[
x = \frac{v - v_f}{v_g - v_f} = \frac{5.0\times 10^{-3} - 1.157\times
10^{-3}}{0.1272 - 1.157\times 10^{-3}} = \frac{3.843\times 10^{-3}}{0.1260}
= 0.0305.
\]
Only $3\,\%$ of the mass is vapour, yet that vapour occupies
$x\, v_g/v = (0.0305)(0.1272)/(5.0\times 10^{-3}) = 0.78$, that is
$78\,\%$ of the vessel volume. The lever rule
(figure~\ref{fig:vol04ch02:quality-lever}) reads this directly: the state
sits very close to the saturated-liquid limb, so the tie-line segment
from $v_f$ to the state is a small fraction of the full tie length, and
the vapour mass fraction is correspondingly small even though the vapour
fills most of the vessel. The asymmetry is the working content of the
lever rule: vapour mass fraction and vapour volume fraction are far apart
whenever $v_g \gg v_f$, which is the rule rather than the exception away
from the critical point. The enthalpy of the charge is $h = (1 -
0.0305)(852.3) + (0.0305)(2792.0) = 911\,\text{kJ}/\text{kg}$, and the
total internal energy follows once $u_f$ and $u_g$ are read from the same
table row.

\subsection{Clausius-Clapeyron extrapolation of a refrigerant vapour
pressure}

The integrated Clausius-Clapeyron relation lets an engineer carry a
single measured saturation point to a nearby temperature without a full
table. We extrapolate the vapour pressure of R-134a from its
$-20\,\degC$ anchor to $0\,\degC$ and check against the table. The anchor
is $p_1 = 132.7\,\text{kPa}$ at $T_1 = 253.15\,\si{\kelvin}$, with a
latent heat in this range of $h_{fg} \approx 212\,\text{kJ}/\text{kg}$
and a specific gas constant $R_s = R/M = 8.314/0.10203 = 81.5\,\text{J}/
(\text{kg}\cdot\si{\kelvin})$. The integrated form
\[
\ln\frac{p_2}{p_1} = -\frac{h_{fg}}{R_s}\left(\frac{1}{T_2} -
\frac{1}{T_1}\right)
\]
with $T_2 = 273.15\,\si{\kelvin}$ gives
\[
\ln\frac{p_2}{p_1} = -\frac{212\times 10^3}{81.5}\left(\frac{1}{273.15}
- \frac{1}{253.15}\right) = -(2601)(-2.893\times 10^{-4}) = 0.752,
\]
so $p_2 = 132.7\, e^{0.752} = 132.7\times 2.121 = 281\,\text{kPa}$. The
R-134a table value at $0\,\degC$ is $292.8\,\text{kPa}$, so the
extrapolation lands within $4\,\%$ over a $20\,\si{\kelvin}$ span. The
small undershoot is the expected sign: the relation assumes a constant
$h_{fg}$ and an ideal vapour with $v_g \gg v_f$, and R-134a near
$0\,\degC$ is a heavier vapour closer to its critical point than steam at
the boiling point, so the straight-line slope of $\ln p$ against $1/T$
drifts. The discipline is to extrapolate only a short distance from a
measured anchor and to expect a few percent of error that grows with the
span. The routine \texttt{code/clausius\_clapeyron.py} carries the same
two-point form.

\subsection{Three cubic equations of state for compressed methane}

A working engineer choosing an equation of state wants to know how much
the choice changes the answer. We put the ideal-gas law, the van der
Waals equation, and the Peng-Robinson equation against one another for
methane at $200\,\si{\kelvin}$ confined to a molar volume of $v =
6.0\times 10^{-4}\,\text{m}^3/\text{mol}$, a compressed state on the way
to liquefaction. Methane has $T_c = 190.6\,\si{\kelvin}$, $p_c =
4.60\,\text{MPa}$, and an acentric factor $\omega = 0.011$, so the
reduced temperature is $T_r = 200/190.6 = 1.05$, just above critical,
where the equations part company.

The ideal-gas pressure is
\[
p_{\text{ideal}} = \frac{RT}{v} = \frac{(8.314)(200)}{6.0\times 10^{-4}}
= 2.77\times 10^6\,\text{Pa} = 2.77\,\text{MPa}.
\]
For van der Waals, $a = 27R^2T_c^2/(64p_c) = 0.230\,\text{Pa}\cdot
\text{m}^6/\text{mol}^2$ and $b = RT_c/(8p_c) = 4.31\times
10^{-5}\,\text{m}^3/\text{mol}$, so
\[
p_{\text{vdW}} = \frac{RT}{v - b} - \frac{a}{v^2}
= \frac{1662.8}{5.569\times 10^{-4}} - \frac{0.230}{3.6\times 10^{-7}}
= 2.986\times 10^6 - 6.39\times 10^5 = 2.35\,\text{MPa}.
\]
For Peng-Robinson the constants are $a = 0.45724\,R^2T_c^2/p_c =
0.249\,\text{Pa}\cdot\text{m}^6/\text{mol}^2$ and $b = 0.07780\,
RT_c/p_c = 2.68\times 10^{-5}\,\text{m}^3/\text{mol}$, and the
temperature function uses $\kappa = 0.37464 + 1.54226\,\omega -
0.26992\,\omega^2 = 0.391$ so that
\[
\alpha(T) = \left[1 + \kappa\left(1 - \sqrt{T_r}\right)\right]^2
= \left[1 + 0.391\left(1 - \sqrt{1.05}\right)\right]^2 = 0.981.
\]
The Peng-Robinson pressure is
\[
p_{\text{PR}} = \frac{RT}{v - b} - \frac{a\,\alpha}{v(v+b) + b(v-b)}.
\]
The repulsive term is $1662.8/(6.0\times 10^{-4} - 2.68\times 10^{-5}) =
2.897\times 10^6\,\text{Pa}$. The attractive denominator is
$v(v+b) + b(v-b) = (6.0\times 10^{-4})(6.268\times 10^{-4}) +
(2.68\times 10^{-5})(5.732\times 10^{-4}) = 3.914\times 10^{-7}$, so the
attractive term is $(0.249)(0.981)/(3.914\times 10^{-7}) = 6.24\times
10^5\,\text{Pa}$. Then $p_{\text{PR}} = 2.897\times 10^6 - 6.24\times
10^5 = 2.27\,\text{MPa}$. The three compressibility factors are
$Z_{\text{ideal}} = 1.00$, $Z_{\text{vdW}} = 0.85$, and $Z_{\text{PR}} =
0.82$. The reference value from a multiparameter methane equation of
state at this state is $Z \approx 0.81$, so Peng-Robinson lands within a
percent, van der Waals within five percent, and the ideal-gas law is
high by more than twenty percent. The ordering is the working lesson: the
ideal-gas law sets the scale, van der Waals captures most of the
correction with two constants and no acentric factor, and Peng-Robinson
earns its place in process software by closing the last few percent that
a custody-transfer or liquefaction calculation cannot give away. The
routine \texttt{code/ideal\_vs\_real\_gas.py} returns the van der Waals
and ideal numbers; the Peng-Robinson arithmetic above is the one a
process simulator runs at every flowsheet node.

\subsection{Throttling a saturated liquid into the two-phase dome}

A throttle, an expansion valve, or a partly-closed restriction holds
enthalpy constant while it drops the pressure, and the question that
matters for a refrigeration or a let-down line is how much of the liquid
flashes to vapour on the way through. We carry saturated-liquid R-134a
from a condenser at $40\,\degC$ ($p = 1017\,\text{kPa}$) through an
expansion valve to an evaporator at $-10\,\degC$ ($p = 201\,\text{kPa}$)
and find the exit state. The throttle is isenthalpic, so the exit
enthalpy equals the inlet saturated-liquid enthalpy,
\[
h_{\text{out}} = h_f(40\,\degC) = 107.3\,\text{kJ}/\text{kg}.
\]
At the exit pressure the saturated-liquid and saturated-vapour
enthalpies are $h_f(-10\,\degC) = 38.6\,\text{kJ}/\text{kg}$ and
$h_g(-10\,\degC) = 244.5\,\text{kJ}/\text{kg}$
(\texttt{data/r134a\_saturation.csv}). Since $h_f < h_{\text{out}} <
h_g$, the exit is a two-phase mixture with quality
\[
x = \frac{h_{\text{out}} - h_f}{h_g - h_f} = \frac{107.3 - 38.6}{244.5 -
38.6} = \frac{68.7}{205.9} = 0.334.
\]
A third of the liquid has flashed to vapour, and the flash is the point
of the throttle: vapourising that third pulls its latent heat out of the
remaining liquid, which is what drops the stream from $40\,\degC$ to
$-10\,\degC$. The cooling per kilogram available downstream is the
evaporator's job, $h_g(-10\,\degC) - h_{\text{out}} = 244.5 - 107.3 =
137.2\,\text{kJ}/\text{kg}$, the remaining two-thirds of the latent heat
that the cold coil can absorb. The throttle is irreversible, and the
entropy rises across it even though the enthalpy does not: the inlet
entropy is $s_f(40\,\degC) = 0.393\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$ and the exit entropy is $s_f(-10\,\degC) + x[s_g(-10\,
\degC) - s_f(-10\,\degC)] = 0.158 + 0.334(0.937 - 0.158) = 0.418\,
\text{kJ}/(\text{kg}\cdot\si{\kelvin})$, an entropy generation of
$0.025\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$ that the expansion valve
trades for the simplicity of having no moving parts. A turbine in place
of the valve would recover some of that lost work, which is why large
refrigeration and liquefaction plants sometimes replace the throttle
with an expander; for a household refrigerator the valve's simplicity
wins.

\subsection{Dew point and the onset of condensation}

Moist air carries water vapour as one component of a gas mixture, and the
water condenses when the air is cooled to the temperature at which its
vapour pressure equals the saturation pressure. We take room air at
$20\,\degC$ and $73\,\%$ relative humidity and find the dew point, the
temperature a window or a cold pipe must reach before it sweats. Relative
humidity is the ratio of the actual water-vapour partial pressure to the
saturation pressure at the air temperature,
\[
\phi = \frac{p_v}{p_{\text{sat}}(T)},
\]
so the partial pressure of water vapour is $p_v = \phi\,
p_{\text{sat}}(20\,\degC) = 0.73\times 2.339 = 1.71\,\text{kPa}$, using
the steam-table saturation pressure of water at $20\,\degC$. Cooling the
air at constant total pressure holds $p_v$ fixed, because the amount of
water in the parcel does not change, and the dew point is the
temperature at which the saturation pressure has fallen to this $p_v$. We
read it off the saturation curve of
figure~\ref{fig:vol04ch02:dewpoint-construction}: the saturation pressure
of water equals $1.71\,\text{kPa}$ at $15\,\degC$, so the dew point is
$15\,\degC$. Any surface in the room colder than $15\,\degC$, a
single-glazed window on a winter night, an uninsulated chilled-water
pipe, the outside of a cold drink, drops the adjacent air below its dew
point and grows condensation. The same construction sets a design rule
for building envelopes: insulation must keep the inner surface of an
exterior wall above the room-air dew point, or water condenses inside the
wall and the assembly rots. The humidity ratio (mass of water vapour per
mass of dry air) follows from the partial pressures through the ideal-gas
law for each component, $w = 0.622\,p_v/(p - p_v) = 0.622\times
1.71/(101.3 - 1.71) = 0.0107$, that is $10.7\,\text{g}$ of water per
kilogram of dry air, the quantity an air-conditioning coil must remove to
dehumidify. The dew-point calculation rests on treating each gas in the
mixture as ideal and additive, the Dalton partial-pressure model, which
holds well for the low water-vapour pressures of atmospheric air and
fails only when a component approaches its own saturation curve at high
pressure.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-dewpoint-construction.tex}

\subsection{Cooling-and-dehumidification load and the sensible-latent
split}

An air-conditioning coil does two jobs at once: it lowers the dry-bulb
temperature of the air (the sensible load) and it condenses water out of it
(the latent load), and the engineer who sizes the coil must split the total
between them, because the two are carried by different equipment limits. We
size the load for a coil treating $0.5\,\text{kg}/\text{s}$ of dry air from
a supply state A at $30\,\degC$ and $50\,\%$ relative humidity to a coil-exit
state B at $14\,\degC$ and $90\,\%$ relative humidity, the process drawn on
the psychrometric skeleton of
figure~\ref{fig:vol04ch02:psychrometric-process}.

We need the humidity ratio at each end. At state A the saturation pressure
of water at $30\,\degC$ is $4.246\,\text{kPa}$, so the vapour partial
pressure is $p_{v,A} = 0.50\times 4.246 = 2.123\,\text{kPa}$ and the
humidity ratio is
\[
w_A = 0.622\,\frac{p_{v,A}}{p - p_{v,A}} = 0.622\times \frac{2.123}{101.3 -
2.123} = 0.0133,
\]
that is $13.3\,\text{g}$ of water per kilogram of dry air. At state B the
saturation pressure at $14\,\degC$ is $1.599\,\text{kPa}$, so $p_{v,B} =
0.90\times 1.599 = 1.439\,\text{kPa}$ and
\[
w_B = 0.622\times \frac{1.439}{101.3 - 1.439} = 0.00896,
\]
about $9.0\,\text{g}/\text{kg}$. The air leaves the coil colder and drier.

The total enthalpy of moist air, per kilogram of dry air, is the sum of the
dry-air and the water-vapour terms, $h = c_{p,a}T + w(h_{g0} + c_{p,v}T)$,
with $c_{p,a} = 1.005\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$, $c_{p,v} =
1.86\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$, and $h_{g0} = 2501\,\text{kJ}/
\text{kg}$ the reference latent heat at $0\,\degC$. At state A,
\[
h_A = (1.005)(30) + 0.0133[2501 + (1.86)(30)] = 30.15 + 0.0133(2556.8) =
64.2\,\text{kJ}/\text{kg},
\]
and at state B,
\[
h_B = (1.005)(14) + 0.00896[2501 + (1.86)(14)] = 14.07 + 0.00896(2527.0) =
36.7\,\text{kJ}/\text{kg}.
\]
The total coil load per kilogram of dry air is $h_A - h_B = 64.2 - 36.7 =
27.5\,\text{kJ}/\text{kg}$, and at $0.5\,\text{kg}/\text{s}$ of dry air the
coil duty is $\dot{Q}_{\text{total}} = (0.5)(27.5) = 13.8\,\text{kW}$.

The split follows by separating the temperature change from the moisture
change. The sensible load is the enthalpy the coil would remove if the
humidity ratio held fixed at its mean,
\[
\dot{Q}_{\text{sens}} = \dot{m}_a\,(c_{p,a} + \bar{w}\,c_{p,v})(T_A - T_B) =
(0.5)(1.005 + 0.0111\times 1.86)(30 - 14) = (0.5)(1.026)(16) =
8.2\,\text{kW},
\]
using the mean humidity ratio $\bar{w} = 0.0111$. The latent load is the
heat carried by the condensed water,
\[
\dot{Q}_{\text{lat}} = \dot{m}_a\,(w_A - w_B)\,h_{fg} = (0.5)(0.0133 -
0.00896)(2470) = (0.5)(0.00434)(2470) = 5.4\,\text{kW},
\]
with $h_{fg} \approx 2470\,\text{kJ}/\text{kg}$ near the coil temperature.
The two sum to $8.2 + 5.4 = 13.6\,\text{kW}$, recovering the total to within
the rounding of the mean-$w$ approximation. The \engterm{sensible heat
ratio} is $\dot{Q}_{\text{sens}}/\dot{Q}_{\text{total}} = 8.2/13.6 = 0.60$,
so $40\,\%$ of this coil's work is dehumidification. The split matters
because the latent load sets the condensate drainage and the coil surface
temperature (the coil must run below the air's dew point of about $18\,\degC$
to condense water at all), while the sensible load sets the air-side
temperature drop; a coil sized on total load alone, with no check that its
surface sits below the dew point, will cool the air without drying it and
leave the space clammy.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-psychrometric-process.tex}

\subsection{An enthalpy departure function from the van der Waals equation}

A property table gives enthalpy directly, but the engineer who works from an
equation of state must build the enthalpy from it, and the bridge is the
\engterm{departure function}: the difference between the real-fluid property
and the ideal-gas property at the same temperature and pressure. We derive
the enthalpy departure for the van der Waals fluid and size it for carbon
dioxide, the substance the transcritical case study leaned on. The general
result from the thermodynamic property relations is
\[
h - h^{\text{ideal}} = \int_{\infty}^{v}\left[T\left(\frac{\partial
p}{\partial T}\right)_v - p\right]dv + pv - RT,
\]
the integral running from the ideal-gas limit ($v\to\infty$) in to the real
molar volume. For the van der Waals equation $p = RT/(v-b) - a/v^2$, the
temperature derivative at constant volume is $(\partial p/\partial T)_v =
R/(v-b)$, so the bracket is
\[
T\left(\frac{\partial p}{\partial T}\right)_v - p = \frac{RT}{v-b} -
\left(\frac{RT}{v-b} - \frac{a}{v^2}\right) = \frac{a}{v^2}.
\]
The integral is elementary,
\[
\int_{\infty}^{v}\frac{a}{v'^2}\,dv' = -\frac{a}{v}\bigg|_{\infty}^{v} =
-\frac{a}{v},
\]
and adding $pv - RT$ gives the closed form
\[
h - h^{\text{ideal}} = -\frac{a}{v} + (pv - RT) = -\frac{a}{v} + RT\!\left(
\frac{v}{v-b} - 1\right) - \frac{a}{v} = \frac{RTb}{v-b} - \frac{2a}{v},
\]
where the second step used $pv = RTv/(v-b) - a/v$. The two terms carry the
two corrections of the equation: the $RTb/(v-b)$ term is the
excluded-volume contribution and raises the enthalpy, the $-2a/v$ term is
the attraction contribution and lowers it. We size it for carbon dioxide at
$310\,\si{\kelvin}$ and $v = 1.5\times 10^{-4}\,\text{m}^3/\text{mol}$, the
compressed state of an earlier worked example, with $a = 0.365\,\text{Pa}
\cdot\text{m}^6/\text{mol}^2$ and $b = 4.28\times 10^{-5}\,\text{m}^3/
\text{mol}$. The excluded-volume term is $RTb/(v-b) = (8.314)(310)(4.28
\times 10^{-5})/(1.072\times 10^{-4}) = 1029\,\text{J}/\text{mol}$, and the
attraction term is $-2a/v = -2(0.365)/(1.5\times 10^{-4}) = -4867\,\text{J}/
\text{mol}$, so
\[
h - h^{\text{ideal}} = 1029 - 4867 = -3838\,\text{J}/\text{mol} =
-3.84\,\text{kJ}/\text{mol}.
\]
Per kilogram, dividing by the molar mass $0.0440\,\text{kg}/\text{mol}$, the
departure is $-87\,\text{kJ}/\text{kg}$: the real compressed carbon dioxide
holds $87\,\text{kJ}/\text{kg}$ less enthalpy than an ideal gas at the same
temperature would, the attraction dominating near the critical region
exactly as the compressibility factor of $0.46$ at this state foretold. An
engineer who priced the enthalpy of this stream from the ideal-gas $c_p\,T$
alone would overstate it by that amount, which is why a custody-transfer or
a compressor-duty calculation on a dense gas runs the departure function
rather than the ideal-gas enthalpy.

\subsection{Latent heat read off the Clausius-Clapeyron slope of two
table points}

The Clausius-Clapeyron relation runs both ways: forward it predicts a
saturation pressure, backward it extracts a latent heat from two measured
saturation points without a calorimeter. We read the latent heat of
ammonia, the industrial-refrigeration working fluid, from its
saturation-pressure table at two temperatures and check it against the
tabulated value. Ammonia saturates at $p_1 = 290.9\,\text{kPa}$ at $T_1 =
-10\,\degC = 263.15\,\si{\kelvin}$ and at $p_2 = 615.2\,\text{kPa}$ at
$T_2 = 10\,\degC = 283.15\,\si{\kelvin}$. The integrated relation
$\ln(p_2/p_1) = -(h_{fg}/R_s)(1/T_2 - 1/T_1)$ solves for the latent heat,
\[
h_{fg} = -R_s\,\frac{\ln(p_2/p_1)}{1/T_2 - 1/T_1},
\]
with the specific gas constant of ammonia $R_s = R/M = 8.314/0.01703 =
488.2\,\text{J}/(\text{kg}\cdot\si{\kelvin})$. The pressure ratio gives
$\ln(615.2/290.9) = \ln(2.115) = 0.749$, and the reciprocal-temperature
difference is $1/283.15 - 1/263.15 = 3.5316\times 10^{-3} - 3.8001\times
10^{-3} = -2.685\times 10^{-4}\,\si{\kelvin}^{-1}$, so
\[
h_{fg} = -(488.2)\,\frac{0.749}{-2.685\times 10^{-4}} = (488.2)(2790) =
1.362\times 10^{6}\,\text{J}/\text{kg} = 1362\,\text{kJ}/\text{kg}.
\]
The tabulated latent heat of ammonia averaged over this range is about
$1240\,\text{kJ}/\text{kg}$, so the two-point slope overstates it by roughly
$10\,\%$. The overshoot is the expected sign and size: the integrated
relation assumes the vapour is ideal and $v_g \gg v_f$, and ammonia at these
pressures is a moderately non-ideal vapour, so the slope of $\ln p$ against
$1/T$ is steeper than the ideal model and the inferred latent heat runs
high. The discipline mirrors the forward extrapolation: a two-point slope is
a quick latent-heat estimate good to about ten percent, and a calorimetric
$h_{fg}$ is the value a cycle calculation should carry. The routine
\texttt{code/clausius\_clapeyron.py} accepts the two points and returns the
same slope.

\subsection{Vapour-pressure and volume change across a Maxwell construction}

The theory section located the saturated states $v_f$ and $v_g$ as the
endpoints of the Maxwell line; here we use them to compute the volume change
of vapourisation that feeds the Clausius-Clapeyron slope, and check that the
two pictures agree. At $T_r = 0.90$ the construction on carbon dioxide
returns a reduced saturation pressure $p_{\text{sat},r} = 0.647$ and reduced
saturation volumes $v_{f,r} = 0.60$ and $v_{g,r} = 2.35$
(figure~\ref{fig:vol04ch02:maxwell-equal-area}). Carbon dioxide has $T_c =
304.1\,\si{\kelvin}$, $p_c = 7.38\,\text{MPa}$, and a critical molar volume
$v_c = 3b = 3(4.28\times 10^{-5}) = 1.284\times 10^{-4}\,\text{m}^3/
\text{mol}$ in the van der Waals model. The dimensional saturation pressure
and volumes at $T = 0.90\,T_c = 273.7\,\si{\kelvin}$ are
\[
p_{\text{sat}} = 0.647\,p_c = 4.78\,\text{MPa}, \quad
v_f = 0.60\,v_c = 7.70\times 10^{-5}, \quad
v_g = 2.35\,v_c = 3.02\times 10^{-4}\,\text{m}^3/\text{mol},
\]
so the volume change of vapourisation is $\Delta v = v_g - v_f = 2.25\times
10^{-4}\,\text{m}^3/\text{mol}$. The Clausius-Clapeyron slope then needs the
latent heat, which the same model supplies through $h_{fg} = T\,\Delta
v\,(dp/dT)$. Estimating the saturation-curve slope from the van der Waals
$p_{\text{sat}}(T)$ near $T_r = 0.90$ as $dp/dT \approx 1.1\times 10^{5}\,
\text{Pa}/\si{\kelvin}$, the molar latent heat is
\[
h_{fg} = T\,\Delta v\,\frac{dp}{dT} = (273.7)(2.25\times 10^{-4})(1.1\times
10^{5}) = 6.78\times 10^{3}\,\text{J}/\text{mol},
\]
about $154\,\text{kJ}/\text{kg}$ after dividing by the molar mass. The
measured latent heat of carbon dioxide at this temperature is roughly
$160\,\text{kJ}/\text{kg}$, so the van der Waals construction lands within
a few percent on the latent heat even though it places the critical
compressibility badly. The lesson is the one the construction promised: the
qualitative geometry of the loop and its equal-area line carries the latent
heat and the volume change correctly to engineering tolerance, while the
absolute volumes near the critical point are where the two-parameter model
shows its limits.

\subsection{The fugacity coefficient of compressed carbon dioxide}

The fugacity is the property a vapour-liquid equilibrium calculation
equates, and the engineer who works from a cubic equation of state computes
it from the compressibility factor. We size the fugacity of carbon dioxide
at the compressed state of the earlier departure example, $310\,\si{\kelvin}$
and $v = 1.5\times 10^{-4}\,\text{m}^3/\text{mol}$, where the van der Waals
equation returned $Z = 0.455$. The cleanest route for the van der Waals
fluid integrates the definition $RT\,d\ln f = v\,dp$ in closed form rather
than the pressure integral; the result, obtained by the same manipulations
that gave the departure functions, is
\[
\ln\phi = \frac{b}{v-b} - \frac{2a}{RTv} - \ln Z,
\]
the three terms carrying the excluded volume, the attraction, and the
$pv$-product correction. With $a = 0.365\,\text{Pa}\cdot\text{m}^6/
\text{mol}^2$, $b = 4.28\times 10^{-5}\,\text{m}^3/\text{mol}$, and $Z =
0.455$, the excluded-volume term is $b/(v-b) = 4.28\times 10^{-5}/
(1.072\times 10^{-4}) = 0.399$, the attraction term is $2a/(RTv) =
2(0.365)/[(8.314)(310)(1.5\times 10^{-4})] = 0.730/0.3866 = 1.889$, and
$\ln Z = \ln 0.455 = -0.787$. Then
\[
\ln\phi = 0.399 - 1.889 - (-0.787) = -0.703, \qquad \phi = e^{-0.703} =
0.495.
\]
The fugacity is $f = \phi\, p = 0.495\times 7.82\,\text{MPa} = 3.87\,
\text{MPa}$, barely half the mechanical pressure. The attraction term
dominates, as the negative enthalpy departure of the same state foretold:
near the critical region carbon dioxide has an escaping tendency far below
its pressure, and a phase-equilibrium calculation that used the pressure in
place of the fugacity would misplace the split between phases. The
generalised chart of figure~\ref{fig:vol04ch02:fugacity-coefficient} reads
$\phi \approx 0.5$ at $T_r = 1.02$ and the corresponding $p_r$, the same
neighbourhood the closed form returns.

\subsection{The entropy departure and the real isentrope of a compressor}

A compressor calculation needs the entropy, not just the enthalpy, because
the isentropic discharge state is found by holding entropy constant, and on
a dense gas the entropy departure is the correction that moves the discharge
off the ideal-gas isentrope. We size the entropy departure for the same
compressed carbon dioxide at $310\,\si{\kelvin}$ and $v = 1.5\times
10^{-4}\,\text{m}^3/\text{mol}$ from the van der Waals result derived in the
property-relations section, $s - s^{\text{ideal}} = R\ln[p(v-b)/RT]$. The
pressure at this state is the van der Waals value $p = 7.82\,\text{MPa}$, so
the argument is
\[
\frac{p(v-b)}{RT} = \frac{(7.82\times 10^6)(1.072\times 10^{-4})}{(8.314)
(310)} = \frac{838.3}{2577.3} = 0.3253,
\]
and the entropy departure is
\[
s - s^{\text{ideal}} = R\ln(0.3253) = (8.314)(-1.123) = -9.34\,\text{J}/
(\text{mol}\cdot\si{\kelvin}),
\]
about $-0.212\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$ after dividing by the
molar mass. The real compressed gas carries less entropy than an ideal gas
at the same temperature and pressure, by the amount its confinement removes.
The working consequence is in the isentropic compression: an ideal-gas model
would predict a discharge temperature from the ideal entropy alone, but the
real isentrope, holding the sum $s^{\text{ideal}}(T,p) + (s -
s^{\text{ideal}})$ constant, lands at a different discharge temperature
because the departure term shifts as the pressure rises through the
compressor. A custody-transfer or a refrigeration-compressor calculation on
carbon dioxide near its critical region runs the entropy departure for
exactly this reason; the ideal-gas isentrope is off by enough to matter for
the discharge temperature and the work.

\subsection{Reading a compressibility factor off the corresponding-states
chart}

The generalised chart turns two reduced coordinates into a compressibility
factor without any substance-specific equation, and the discipline is to
form the reduced coordinates, read $Z$, and check the answer against a
cubic. We size the mass of ethane in a storage bottle this way. A
$0.040\,\text{m}^3$ bottle holds ethane at $290\,\si{\kelvin}$ and
$6.0\,\text{MPa}$; ethane has $T_c = 305.3\,\si{\kelvin}$ and $p_c =
4.87\,\text{MPa}$ (\texttt{data/critical\_constants.csv}). The reduced
coordinates are
\[
T_r = \frac{290}{305.3} = 0.95, \qquad p_r = \frac{6.0}{4.87} = 1.23,
\]
a point below the critical temperature and above the critical pressure,
deep in the region where the chart dips hardest. Reading the generalised
compressibility chart of figure~\ref{fig:vol04ch02:compressibility-chart}
at $T_r = 0.95$ and $p_r = 1.23$ gives $Z \approx 0.40$, a compressed gas
near its saturation curve. The real mass follows from $pV = Z m R_s T$ with
$R_s = R/M = 8.314/0.03007 = 276.5\,\text{J}/(\text{kg}\cdot\si{\kelvin})$,
\[
m = \frac{pV}{Z R_s T} = \frac{(6.0\times 10^6)(0.040)}{(0.40)(276.5)(290)}
= \frac{2.40\times 10^5}{3.207\times 10^4} = 7.5\,\text{kg}.
\]
An ideal-gas estimate would have set $Z = 1$ and returned $3.0\,\text{kg}$,
understating the charge by a factor of $2.5$, because at this reduced state
the gas is two and a half times denser than ideal. The check against a
Peng-Robinson solve at the same state returns $Z \approx 0.38$, within the
reading error of the chart, so the chart's $7.5\,\text{kg}$ stands. The
lesson is the working one: a bottle filled to a measured pressure holds far
more dense gas than the ideal-gas law predicts whenever the state sits near
the saturation curve, and the corresponding-states chart catches the
difference with nothing more than the critical constants.

\subsection{Locating the saturation pressure by equality of fugacity}

The Maxwell equal-area construction and the equality of fugacity are two
faces of the same condition, and putting them side by side fixes both the
geometry and the algebra in the reader's hand. We locate the saturation
pressure of the van der Waals fluid at $T_r = 0.90$ by the fugacity
condition and check it against the equal-area value the chapter already
quoted. Equilibrium requires equal fugacity in the two phases, $f_f = f_g$,
hence equal fugacity coefficient times pressure; at a common saturation
pressure the pressures cancel and the condition reduces to $\ln\phi_f =
\ln\phi_g$, with each coefficient evaluated from the van der Waals closed
form $\ln\phi = b/(v-b) - 2a/(RTv) - \ln Z$ at the liquid root $v_f$ and the
vapour root $v_g$ of the equation at the trial pressure. The procedure is to
guess a saturation pressure, solve the cubic for its three roots, keep the
smallest ($v_f$) and the largest ($v_g$), evaluate $\ln\phi$ at each, and
adjust the pressure until the two agree. At $T_r = 0.90$ this converges to a
reduced saturation pressure $p_{\text{sat},r} = 0.647$ with reduced volumes
$v_{f,r} = 0.60$ and $v_{g,r} = 2.35$, the identical values the equal-area
bisection of \texttt{code/vdw\_isotherms.py} returned in
figure~\ref{fig:vol04ch02:maxwell-equal-area}. The two routes agree because
they encode the same physics: the equal-area construction is the graphical
statement that $\int v\,dp = 0$ around the loop, and the equal-fugacity
condition is the same integral written as $\ln(f_g/f_f) = (1/RT)\int v\,dp =
0$. An engineer using a process simulator never sees the equal-area picture;
the simulator equates fugacities. An engineer reading a $p$-$v$ diagram
never sees the fugacity; the engineer balances the lobes. Both arrive at the
same saturation line, and knowing they are the same condition is what lets a
reader move between the picture and the calculation.

\subsection{The fusion curve, its slope, and the melting-point anomaly of
water}

The chapter's Clausius-Clapeyron work has stayed on the vapourisation curve,
where the vapour dwarfs the liquid and the slope is gentle. The same
relation governs the solid-liquid fusion curve, and there the arithmetic
turns instructive because the volume change on melting is a thousand times
smaller than on boiling, so the slope is a thousand times steeper and its
sign carries a physical surprise. We compute the fusion-curve slope of water
and read off why pressure lowers the melting point of ice. The Clapeyron
relation on the fusion curve is
\[
\frac{dp}{dT} = \frac{h_{sf}}{T\,(v_{\text{liq}} - v_{\text{sol}})},
\]
with $h_{sf} = 333.5\,\text{kJ}/\text{kg}$ the latent heat of fusion of
water at $0\,\degC$. The specific volumes are $v_{\text{liq}} = 1.000\times
10^{-3}\,\text{m}^3/\text{kg}$ for liquid water and $v_{\text{sol}} =
1.091\times 10^{-3}\,\text{m}^3/\text{kg}$ for ice, so the volume change on
melting is $\Delta v = v_{\text{liq}} - v_{\text{sol}} = -0.091\times
10^{-3}\,\text{m}^3/\text{kg}$, negative because ice is the rare solid that
floats on its own melt. The slope is
\[
\frac{dp}{dT} = \frac{333.5\times 10^3}{(273.15)(-0.091\times 10^{-3})} =
-1.34\times 10^7\,\text{Pa}/\si{\kelvin} = -134\,\text{bar}/\si{\kelvin}.
\]
The sign is the anomaly: for water the fusion curve leans backward,
so raising the pressure lowers the melting point, and the magnitude says how
little it moves. To depress the melting point by one degree takes
$134\,\text{bar}$, so the hundred-atmosphere pressure under an ice-skate
blade buys only about $0.75\,\si{\kelvin}$ of melting-point depression, far
too little to account for the slick film under a skate, which is mostly
frictional heating instead. For a normal substance, where the solid is
denser than the liquid and $\Delta v > 0$, the same relation gives a steep
positive slope: the fusion curve leans forward, and pressure raises the
melting point. Figure~\ref{fig:vol04ch02:fusion-curve-slope} sets the two
cases side by side against the gentle vapourisation curve. The working
lesson is that the Clapeyron relation reads the microscopic packing off a
macroscopic slope: the sign of the fusion-curve slope is the sign of the
density change on freezing, and water's backward-leaning curve is the direct
thermodynamic statement that ice is less dense than water.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-fusion-curve-slope.tex}

\subsection{The speed of sound as an equation-of-state derivative}

A pressure disturbance too fast for heat to diffuse propagates at the speed
of sound, and that speed is an equation-of-state derivative, the isentropic
stiffness of the fluid, which ties the acoustic behaviour of a gas back to
its $p$-$v$-$T$ surface. The speed of sound is
\[
a = \sqrt{\left(\frac{\partial p}{\partial \rho}\right)_s},
\]
the isentropic derivative of pressure with density. For an ideal gas an
isentropic process obeys $p\rho^{-\gamma} = \text{const}$, so $(\partial
p/\partial \rho)_s = \gamma p/\rho = \gamma R_s T$, and the speed of sound
is $a = \sqrt{\gamma R_s T}$, depending on temperature alone. For air at
$20\,\degC$ with $\gamma = 1.4$ and $R_s = 287\,\text{J}/(\text{kg}\cdot
\si{\kelvin})$,
\[
a = \sqrt{(1.4)(287)(293.15)} = 343\,\text{m}/\text{s},
\]
the familiar figure. For superheated steam near a turbine inlet at
$400\,\degC$, with $\gamma \approx 1.3$ and $R_s = 462\,\text{J}/(\text{kg}
\cdot\si{\kelvin})$, the same formula gives $a = \sqrt{(1.3)(462)(673)} =
636\,\text{m}/\text{s}$, the lighter and hotter molecules carrying sound
faster. Two consequences follow for the working engineer. The first is
choking: a nozzle or a relief orifice passing gas cannot accelerate the flow
past the local sound speed at its throat, so once the pressure ratio is
large enough to reach $\text{Mach}\ 1$ the mass flow saturates, which is why
the relief-valve sizing of the case study depends on the sonic vapour
properties at the set pressure and not on the downstream pressure. The
second is the contrast with the liquid: the incompressibility section found
a water sound speed of $1480\,\text{m}/\text{s}$ from the bulk modulus,
$a = \sqrt{K/\rho}$, four times the air value, because the liquid is far
stiffer. The two-phase mixture is stranger still. A bubbly liquid combines
the high density of the liquid with the high compressibility of the vapour,
and its sound speed drops below that of either pure phase, to as little as
$20$ to $30\,\text{m}/\text{s}$ for an air-water froth, which is why a
cavitating pump or a steam-water line carries acoustic transients so
different from the single-phase lines around it.

\subsection{An isothermal flash of a binary mixture}

Every worked example so far has treated a pure substance, but process
streams are mixtures, and the move that splits a mixture between a liquid and
a vapour is the \engterm{isothermal flash}. We carry a binary feed through
one and read the phase split off the equilibrium ratios. A feed of overall
composition $z_1 = 0.35$ in the volatile component (the rest, $z_2 = 0.65$,
the heavy component) enters a flash drum held at a temperature and pressure
where the two components have equilibrium ratios, or K-values, $K_1 = y_1/x_1
= 2.5$ and $K_2 = y_2/x_2 = 0.45$, the ratio of vapour to liquid mole
fraction that a property package returns from the fugacity equality of the
fugacity section. The K-value above one means the light component prefers the
vapour, the K-value below one means the heavy component prefers the liquid.
The vapour fraction $V$ (moles of vapour per mole of feed) solves the
\engterm{Rachford-Rice equation}, the material balance written so the two
mole-fraction sums both close,
\[
\sum_i \frac{z_i\,(K_i - 1)}{1 + V\,(K_i - 1)} = 0.
\]
For the binary this is
\[
\frac{(0.35)(2.5 - 1)}{1 + V(2.5 - 1)} + \frac{(0.65)(0.45 - 1)}{1 +
V(0.45 - 1)} = \frac{0.525}{1 + 1.5V} - \frac{0.3575}{1 - 0.55V} = 0,
\]
which rearranges to a quadratic in $V$ and solves to $V = 0.203$: about a
fifth of the feed leaves as vapour. The phase compositions follow from
$x_i = z_i/[1 + V(K_i - 1)]$ and $y_i = K_i x_i$,
\[
x_1 = \frac{0.35}{1 + (0.203)(1.5)} = 0.268, \quad
y_1 = (2.5)(0.268) = 0.671,
\]
with $x_2 = 0.732$ and $y_2 = 0.329$ closing both sums to one. The vapour is
enriched to $67\,\%$ in the light component from a feed of $35\,\%$, the
enrichment that a distillation column stacks stage on stage. The same split
is the horizontal tie line of the $P$-$x$-$y$ diagram in
figure~\ref{fig:vol04ch02:binary-pxy}: the feed composition sits between the
bubble and dew lines, and the flash lands its liquid on the bubble line and
its vapour on the dew line. The working content is that a single drum, given
a temperature, a pressure, and the K-values the equation of state supplies,
divides any feed into two streams of definite composition, and the
Rachford-Rice root is the one arithmetic every flowsheet solves at every
flash node.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-binary-pxy.tex}

\subsection{Counting degrees of freedom with the Gibbs phase rule}

Before an engineer fixes a state, it helps to know how many properties are
free to fix, and the count is the \engterm{Gibbs phase rule},
\[
F = C - P + 2,
\]
with $C$ the number of independent chemical components, $P$ the number of
phases present, and $F$ the number of intensive properties that can be varied
independently without changing how many phases coexist. The rule explains
several facts the chapter has used without naming. For a pure substance
($C = 1$) in a single phase ($P = 1$), $F = 1 - 1 + 2 = 2$: two independent
intensive properties fix the state, which is the principle stated earlier
that $(T, p)$ or $(p, h)$ specifies a single-phase point. For the same pure
substance with two phases coexisting ($P = 2$), $F = 1 - 2 + 2 = 1$: only
one property is free, which is why the saturation pressure is a function of
temperature alone and why a two-phase mixture needs the quality $x$ (an
extensive split, not an intensive property) as a second coordinate. For a
pure substance with three phases ($P = 3$), $F = 1 - 3 + 2 = 0$: nothing is
free, which is exactly why the triple point of water sits at one fixed
temperature and pressure ($0.01\,\degC$, $611\,\text{Pa}$) and served as a
defining fixed point of the temperature scale. Extend to the binary flash:
two components ($C = 2$) in two phases ($P = 2$) give $F = 2 - 2 + 2 = 2$, so
fixing the temperature and the pressure of the flash drum, as the flash
example did, leaves no further freedom and the K-values and phase
compositions are then determined. The rule is the bookkeeping that tells an
engineer, before any calculation, how many knobs a phase-equilibrium problem
actually has, and it catches the common error of trying to fix a
two-phase pure substance by temperature and pressure together, which are not
independent there.

\subsection{The limit of superheat from the van der Waals spinodal}

The metastable region between the saturation line and the stability limit is
where a liquid can be carried above its boiling point without boiling, and
the far edge of that region, the \engterm{limit of superheat}, is the state
past which no liquid can persist. We locate it for water from the van der
Waals spinodal and check the answer against the measured value, because the
limit of superheat is the physical trigger the BLEVE mechanism turns
dangerous. The saturation line, the binodal, is where equilibrium boiling
would begin if a bubble could nucleate; the spinodal, the locus of
$(\partial p/\partial v)_T = 0$, is where the homogeneous liquid loses even
mechanical stability and must flash. Figure~\ref{fig:vol04ch02:spinodal-superheat}
draws both curves in reduced coordinates. The van der Waals spinodal in
reduced variables is the parametric curve
\[
T_r = \frac{(3v_r - 1)^2}{4\,v_r^3}, \qquad p_r = \frac{8T_r}{3v_r - 1} -
\frac{3}{v_r^2},
\]
found by setting the reduced isotherm's volume derivative to zero. For water
at atmospheric pressure the reduced pressure is $p_r = p_{\text{atm}}/p_c =
101.3\,\text{kPa}/22.06\,\text{MPa} = 0.0046$, deep in the low-pressure
region. Solving the spinodal parametrisation for the liquid branch (the
small-$v_r$ root) at this $p_r$ returns $T_r = 0.844$, so the limit of
superheat is
\[
T_{\text{sl}} = T_r\,T_c = (0.844)(647.1\,\si{\kelvin}) = 546\,\si{\kelvin}
= 273\,\degC.
\]
Water at atmospheric pressure can in principle be carried to $273\,\degC$ as
a liquid before it must flash, more than $170\,\si{\kelvin}$ above its
$100\,\degC$ boiling point. The measured limit of superheat of water, from
bubble-column and pulse-heating experiments, is about $270\,\degC$ to
$280\,\degC$, so the crude van der Waals spinodal lands within a few degrees
of it, one of the model's better quantitative predictions. The engineering
weight of the number is in the failure section: a pressurised liquid held
well above its atmospheric boiling point sits in exactly this superheated
region, and when a vessel wall tears and the pressure drops to atmospheric,
the liquid finds itself past the atmospheric spinodal, where no metastable
state remains, and it flashes explosively rather than gently. The limit of
superheat is the reason a BLEVE is a flash and not a boil, and the spinodal
is where the chapter's stability archetype meets its most violent failure
mode.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-spinodal-superheat.tex}

\subsection{A vignette: the dry adiabatic lapse rate and the cloud base}

The equation of state and the saturation curve act together to set the
height at which a cloud forms, and the calculation is a clean
two-substance application of the chapter's tools to the atmosphere. A parcel
of air pushed upward, by a hillside or by surface heating, expands as it
rises into lower pressure and cools. The expansion is fast enough and the
air a poor enough conductor that the parcel exchanges almost no heat with
its surroundings, so the process is adiabatic, and for an adiabatic
ideal-gas process the temperature falls with pressure. Combining the
hydrostatic pressure gradient $dp/dz = -\rho g$ with the dry-adiabatic
relation gives the \engterm{dry adiabatic lapse rate},
\[
\Gamma_d = -\frac{dT}{dz} = \frac{g}{c_p},
\]
the temperature drop per unit height of an unsaturated parcel. With $g =
9.81\,\text{m}/\text{s}^2$ and the specific heat of air $c_p =
1005\,\text{J}/(\text{kg}\cdot\si{\kelvin})$, the rate is $\Gamma_d =
9.81/1005 = 9.76\times 10^{-3}\,\si{\kelvin}/\text{m}$, that is $9.8\,
\si{\kelvin}$ per kilometre, the figure drawn in
figure~\ref{fig:vol04ch02:lapse-rate-cloud-base}. The derivation uses only
the ideal-gas law for dry air, the first law with no heat exchange, and the
hydrostatic balance.

The cloud forms where the rising parcel reaches saturation, and finding that
height is the dew-point construction of the earlier worked example carried up
through the atmosphere. The parcel starts at the surface with a temperature
and a dew point; we take a $25\,\degC$ surface at a dew point of $12\,\degC$,
an ordinary humid afternoon. As the parcel rises its temperature falls along
the dry adiabat at $9.8\,\si{\kelvin}/\text{km}$, while its dew point, the
saturation temperature of the water it carries, falls much more slowly,
because lifting at constant water content barely changes the vapour's
partial pressure and the Clausius-Clapeyron saturation curve is steep; the
dew point drops at only about $1.8\,\si{\kelvin}/\text{km}$. The two
temperatures close on each other at the combined rate of $9.8 - 1.8 =
8.0\,\si{\kelvin}/\text{km}$, and the parcel reaches saturation when the gap
of $25 - 12 = 13\,\si{\kelvin}$ has closed,
\[
z_{\text{LCL}} = \frac{T_{\text{surface}} - T_{\text{dew}}}{\Gamma_d -
\Gamma_{\text{dew}}} = \frac{13}{8.0\,\si{\kelvin}/\text{km}} =
1.6\,\text{km},
\]
the \engterm{lifting condensation level}, the flat base a fair-weather cumulus
cloud shows on a summer day. The rule of thumb that meteorologists carry,
that the cloud base in metres is about $125$ times the surface
temperature-dew-point spread in degrees Celsius, is this same result
arithmetically rounded: $125\times 13 = 1625\,\text{m}$.

Above the cloud base the parcel is saturated, and further lifting condenses
water, which releases latent heat back into the parcel and slows its
cooling. The parcel now follows the gentler \engterm{moist adiabatic lapse
rate}, roughly $6\,\si{\kelvin}/\text{km}$ in the lower troposphere, the
amber line of the figure; the rate is not constant because the latent-heat
release depends on how much water is left to condense, which falls as the
parcel cools and dries. The whole construction is the chapter in miniature:
the ideal-gas law and the first law set the dry lapse rate, the
Clausius-Clapeyron saturation curve sets the dew-point lapse rate and the
condensation height, and the latent heat of vapourisation sets the moist
lapse rate above the cloud base. An engineer sizing a cooling-tower plume, a
ventilation stack, or an aircraft icing envelope runs the same three pieces;
the atmosphere is a working fluid with an equation of state and a saturation
curve, and it obeys the chapter like any other.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-lapse-rate-cloud-base.tex}

\subsection{A vignette: the pressure cooker and the autoclave}

A sealed pot raises the boiling point of water by holding the vapour above
it at more than atmospheric pressure, and the same move sterilises a surgical
tray. The device is a clean single-substance application of the saturation
curve, and it repays a close reading because the whole of its usefulness
sits in one point on the curve. A pot open to the air boils water at
$100\,\degC$ because the saturation pressure of water reaches atmospheric at
that temperature, and no amount of extra heat lifts the temperature further:
the added heat vapourises water at constant temperature rather than warming
it, so an open pot is pinned to $100\,\degC$ as long as liquid remains. A
pressure cooker seals the vapour space and lets the pressure rise until a
weighted valve lifts. A domestic cooker vents at a gauge pressure of about
one atmosphere, so the absolute pressure inside is near $200\,\text{kPa}$,
and the water boils at the saturation temperature for that pressure. Reading
the saturation curve, or carrying the Clausius-Clapeyron relation forward
from the $100\,\degC$ anchor,
\[
\ln\frac{p_2}{p_1} = -\frac{h_{fg}}{R_s}\left(\frac{1}{T_2} -
\frac{1}{T_1}\right),
\]
with $p_2 = 198.7\,\text{kPa}$, $p_1 = 101.3\,\text{kPa}$, $h_{fg} =
2257\,\text{kJ}/\text{kg}$, and $R_s = 461.5\,\text{J}/(\text{kg}\cdot
\si{\kelvin})$, returns a boiling temperature of $120\,\degC$. The twenty
degrees are the whole point. Cooking rates, like most chemical rates, climb
steeply with temperature, roughly doubling to tripling for every
$10\,\si{\kelvin}$, so a twenty-degree lift speeds the softening of tough
tissue and the gelatinisation of starches by a factor of four to nine, which
is why a pressure cooker turns a three-hour braise into a forty-minute one.
The valve, not the heat setting, fixes the temperature: once the valve is
lifting, the cooker holds the $200\,\text{kPa}$ saturation state, and turning
the burner higher only vents faster without raising the temperature, the
same constant-temperature pinning an open pot shows, moved up the saturation
curve by the sealed pressure.

The autoclave is the pressure cooker taken to the specification of a
standard. Steam sterilisation runs at a gauge pressure near
$103\,\text{kPa}$ (about $15\,\text{psi}$), an absolute pressure of roughly
$205\,\text{kPa}$, at which saturated steam sits at $121\,\degC$, the
canonical autoclave temperature. The choice of $121\,\degC$ is not
arbitrary: microbial death is a rate process with its own temperature
dependence, characterised by a decimal-reduction time that falls by a factor
of ten for roughly every $10\,\si{\kelvin}$ of temperature, and $121\,\degC$
held for fifteen minutes delivers a reliable kill of bacterial spores that
survive boiling at $100\,\degC$ for hours. The working fluid must be
saturated steam and not hot air, because the sterilising power comes from the
latent heat the condensing steam dumps onto the cold load: when steam at
$121\,\degC$ meets a $20\,\degC$ instrument tray it condenses, releasing
$2200\,\text{kJ}/\text{kg}$ into the surface and driving it rapidly to the
steam temperature, a heat transfer no dry gas at the same temperature can
match. An autoclave that traps air pockets, or that runs on superheated
rather than saturated steam, sterilises unreliably because the load never
reaches the saturation temperature the condensing latent heat would
otherwise guarantee, which is why autoclave cycles begin with an
air-evacuation or steam-flush phase. The two devices are one idea read at
two points on the saturation curve: seal the vapour space, let the pressure
rise to a set point, and the liquid and its steam sit at the single
saturation temperature that point commands, higher than the open-air boiling
point by exactly the amount the Clausius-Clapeyron slope prescribes.

\subsection{The Boyle temperature from the second virial coefficient}

The virial expansion writes the compressibility factor as a power series in
density, and its leading coefficient carries a temperature at which a real
gas mimics an ideal one for a reason that has nothing to do with being
dilute. We locate that temperature for nitrogen from the van der Waals second
virial coefficient and read off what it means. Expanding the van der Waals
pressure $p = RT/(v-b) - a/v^2$ in powers of $1/v$, and using $1/(v-b) =
(1/v)(1 + b/v + b^2/v^2 + \cdots)$, gives
\[
Z = \frac{pv}{RT} = 1 + \frac{1}{v}\left(b - \frac{a}{RT}\right) +
\frac{b^2}{v^2} + \cdots,
\]
so the second virial coefficient is $B(T) = b - a/(RT)$. It is negative at
low temperature, where the attraction term $a/(RT)$ dominates and the gas is
easier to compress than ideal, and positive at high temperature, where the
excluded volume $b$ dominates and the gas is harder to compress than ideal.
The crossing, where $B(T) = 0$, is the \engterm{Boyle temperature}
\[
T_B = \frac{a}{Rb}.
\]
For nitrogen the van der Waals constants are $a = 0.137\,\text{Pa}\cdot
\text{m}^6/\text{mol}^2$ and $b = 3.87\times 10^{-5}\,\text{m}^3/\text{mol}$,
so
\[
T_B = \frac{0.137}{(8.314)(3.87\times 10^{-5})} = \frac{0.137}{3.218\times
10^{-4}} = 426\,\si{\kelvin}.
\]
The measured Boyle temperature of nitrogen is $327\,\si{\kelvin}$, so the van
der Waals estimate runs high by the familiar margin, but it lands in the
right neighbourhood and gives the right physical picture, drawn in reduced
form in figure~\ref{fig:vol04ch02:boyle-virial-B}. In terms of the critical
temperature the same relation reads $T_B = a/(Rb) = (27/8)\,T_c = 3.375\,T_c$,
which for nitrogen's $T_c = 126.2\,\si{\kelvin}$ returns $426\,\si{\kelvin}$
directly and shows the Boyle temperature to be a fixed multiple of the
critical temperature in the van der Waals model. The working content is that
a gas held near its Boyle temperature obeys $pv = RT$ to first order in
density over a wide pressure band, so a nitrogen measurement near
$330\,\si{\kelvin}$ behaves more ideally than the reduced pressure alone
would suggest, because the two corrections cancel rather than vanish. An
engineer calibrating a gas thermometer or a flow meter chooses the working
temperature near $T_B$ for exactly this cancellation.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-boyle-virial-B.tex}

\subsection{Joule-Thomson cooling and the sign of a throttled leak}

A throttle across a single-phase gas holds enthalpy constant and changes the
temperature, and the sign of that change decides whether a high-pressure leak
frosts or self-heats. We size the Joule-Thomson coefficient for carbon
dioxide at room conditions and check its sign against the inversion curve.
The coefficient is
\[
\mu_{JT} = \left(\frac{\partial T}{\partial p}\right)_h = \frac{1}{c_p}
\left[T\left(\frac{\partial v}{\partial T}\right)_p - v\right],
\]
and for a van der Waals gas at low to moderate density the bracket reduces, to
leading order, to $T(\partial v/\partial T)_p - v \approx (2a/RT) - b$, so
\[
\mu_{JT} \approx \frac{1}{c_p}\left(\frac{2a}{RT} - b\right).
\]
For carbon dioxide $a = 0.365\,\text{Pa}\cdot\text{m}^6/\text{mol}^2$, $b =
4.28\times 10^{-5}\,\text{m}^3/\text{mol}$, and the molar heat capacity is
$c_p \approx 37\,\text{J}/(\text{mol}\cdot\si{\kelvin})$. At $T = 298\,
\si{\kelvin}$ the attraction group is $2a/(RT) = 2(0.365)/[(8.314)(298)] =
0.730/2477.6 = 2.95\times 10^{-4}\,\text{m}^3/\text{mol}$, which exceeds $b =
0.428\times 10^{-4}$, so the bracket is positive and
\[
\mu_{JT} \approx \frac{2.95\times 10^{-4} - 0.428\times 10^{-4}}{37} =
\frac{2.52\times 10^{-4}}{37} = 6.8\times 10^{-6}\,\si{\kelvin}/\text{Pa}
= 0.68\,\si{\kelvin}/\text{bar}.
\]
A carbon-dioxide cylinder throttled from $60\,\text{bar}$ to atmospheric
cools by roughly $\mu_{JT}\,\Delta p \approx (0.68)(59) = 40\,\si{\kelvin}$,
which is why a discharging CO\textsubscript{2} fire extinguisher frosts its
horn and can deposit dry-ice snow. The sign flips where the bracket vanishes,
at the low-pressure inversion temperature $T = 2a/(Rb)$, which for carbon
dioxide is $2(0.365)/[(8.314)(4.28\times 10^{-5})] = 2051\,\si{\kelvin}$,
far above room temperature, so carbon dioxide sits deep inside its inversion
curve at ambient and always cools on throttling. Hydrogen is the opposite
case its low inversion temperature makes dangerous: with the same estimate,
$T_{\text{inv}}^{\max} = 2a/(Rb) \approx 220\,\si{\kelvin}$, below room
temperature, so a room-temperature hydrogen jet leaving a high-pressure leak
sits outside its inversion curve and \emph{warms} as it expands, one of the
mechanisms behind self-ignition of high-pressure hydrogen releases.
Figure~\ref{fig:vol04ch02:joule-thomson-inversion} places both cases on the
inversion curve. The lesson for a designer sizing a let-down station is the
one the coefficient's sign carries: know which side of the inversion curve
the gas sits on before assuming a throttle cools it.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-joule-thomson-inversion.tex}

\subsection{Choked flow and the sizing of a relief orifice}

The speed-of-sound example found that a gas nozzle cannot accelerate its flow
past the local sound speed at the throat, and the consequence, choking, sets
the mass rate a relief valve passes. We carry the arithmetic that turns a
required relief mass rate into an orifice area for the propane fire case. When
the ratio of upstream stagnation pressure $p_0$ to back pressure exceeds the
critical value, the throat is sonic and the mass flux is the maximum
\[
\frac{\dot{m}}{A} = p_0\sqrt{\frac{\gamma}{R_s T_0}}\left(\frac{2}{\gamma +
1}\right)^{(\gamma + 1)/[2(\gamma - 1)]},
\]
a function of the upstream state and the ratio of specific heats alone, not
of the downstream pressure. The critical pressure ratio at which the throat
first reaches Mach one is $p^*/p_0 = [2/(\gamma + 1)]^{\gamma/(\gamma - 1)}$,
which for propane vapour with $\gamma \approx 1.13$ is $0.58$, and for a
diatomic gas with $\gamma = 1.4$ is $0.528$;
figure~\ref{fig:vol04ch02:nozzle-choked-flow} plots the flux rising to this
ratio and then holding flat. The propane relief case demanded a vapour rate
of $\dot{m}_{\text{relief}} = 0.45\,\text{kg}/\text{s}$ leaving at the set
pressure of $p_0 = 2.07\,\text{MPa}$ and a saturation temperature near
$T_0 = 331\,\si{\kelvin}$, with $R_s = 188.5\,\text{J}/(\text{kg}\cdot
\si{\kelvin})$ for propane. The mass flux at the choked throat is
\[
\frac{\dot{m}}{A} = (2.07\times 10^6)\sqrt{\frac{1.13}{(188.5)(331)}}\left(
\frac{2}{2.13}\right)^{2.13/0.26}.
\]
The square-root group is $\sqrt{1.13/62{,}394} = \sqrt{1.811\times 10^{-5}} =
4.256\times 10^{-3}$, and the bracket is $(0.939)^{8.19} = 0.596$, so
\[
\frac{\dot{m}}{A} = (2.07\times 10^6)(4.256\times 10^{-3})(0.596) = 5.25
\times 10^3\,\text{kg}/(\text{m}^2\cdot\text{s}).
\]
The required orifice area is then
\[
A = \frac{\dot{m}_{\text{relief}}}{\dot{m}/A} = \frac{0.45}{5.25\times 10^3}
= 8.6\times 10^{-5}\,\text{m}^2 = 86\,\text{mm}^2,
\]
a bore of about $10\,\text{mm}$ before the discharge coefficient (roughly
$0.85$ for a relief nozzle) enlarges it to near $12\,\text{mm}$. The working
lesson is that the choked-flow relation fixes the relief area from the set
pressure and the vapour properties without any reference to the downstream
piping, which is why relief-valve sizing standards tabulate the mass flux per
unit area at the set pressure and leave the tail-pipe out of the capacity
calculation. A valve sized on the mistaken belief that the downstream
pressure throttles the flow would be undersized, because the sonic throat has
already fixed the rate.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-nozzle-choked-flow.tex}

\subsection{A bubble-point pressure from Raoult's law}

The binary flash equated fugacities to split a feed, and for a mixture whose
components are chemically similar the fugacity equality reduces to a rule
simple enough to carry by hand, \engterm{Raoult's law}, which sets each
component's vapour partial pressure equal to its liquid mole fraction times
its pure-component saturation pressure. We find the bubble-point pressure of
a benzene-toluene liquid and read the vapour it first releases. A liquid of
$40\,\%$ benzene ($x_B = 0.40$, $x_T = 0.60$) is held at $95\,\degC$, where
the pure-component saturation pressures are $p_B^{\text{sat}} = 156\,\text{kPa}$
for benzene and $p_T^{\text{sat}} = 63\,\text{kPa}$ for toluene, the more and
less volatile components. Raoult's law gives the total pressure at which the
liquid is at its bubble point,
\[
p_{\text{bubble}} = x_B\,p_B^{\text{sat}} + x_T\,p_T^{\text{sat}} =
(0.40)(156) + (0.60)(63) = 62.4 + 37.8 = 100.2\,\text{kPa},
\]
just above one atmosphere, so this liquid boils near $95\,\degC$ at ambient
pressure. The composition of the first bubble follows from each component's
partial pressure divided by the total,
\[
y_B = \frac{x_B\,p_B^{\text{sat}}}{p_{\text{bubble}}} = \frac{62.4}{100.2} =
0.623, \qquad y_T = \frac{37.8}{100.2} = 0.377,
\]
so the vapour is enriched from $40\,\%$ benzene in the liquid to $62\,\%$
benzene in the vapour, the enrichment the $T$-$x$-$y$ diagram of
figure~\ref{fig:vol04ch02:txy-distillation} shows as the horizontal gap
between the bubble and dew lines. The relative volatility, the ratio of the
two K-values, is $\alpha = (y_B/x_B)/(y_T/x_T) = (156/63) = 2.48$, and it is
this number, larger than one, that makes the separation possible: were
$\alpha = 1$ the vapour and liquid would share a composition and no
distillation could enrich the stream. The working content is that a
similar-molecule mixture needs no equation-of-state solve for a first pass,
because Raoult's law hands the bubble-point pressure and the vapour
composition from the pure-component saturation curves alone, and the departure
from Raoult's law is the correction a fugacity-based flash restores for
dissimilar molecules.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-txy-distillation.tex}

\subsection{The adiabatic flame-free temperature rise of a compressed gas}

An isentropic compression raises a gas's temperature, and the rise is an
equation-of-state result that sizes the discharge temperature of a compressor
and the auto-ignition risk of a diesel or a fast-filled cylinder. We compute
it for air and for a real-gas correction. For an ideal gas the isentropic
relation $Tp^{(1-\gamma)/\gamma} = \text{const}$ gives the discharge
temperature from the pressure ratio,
\[
\frac{T_2}{T_1} = \left(\frac{p_2}{p_1}\right)^{(\gamma - 1)/\gamma}.
\]
Air drawn at $T_1 = 300\,\si{\kelvin}$ and compressed through a pressure
ratio of $20$ (a diesel compression ratio expressed in pressure) with
$\gamma = 1.4$ reaches
\[
T_2 = 300\times 20^{0.2857} = 300\times 2.354 = 706\,\si{\kelvin} =
433\,\degC,
\]
above the auto-ignition temperature of diesel fuel, which is the whole
principle of compression ignition: the charge lights on the heat of
compression alone, with no spark. A fast-filled scuba or breathing-air
cylinder shows the same rise on a smaller pressure ratio and is the reason
rapid filling warms the bottle and, if oil contamination is present, risks a
compression-ignition fire. The real-gas correction enters through the
temperature dependence of $\gamma$ and through the departure functions: at
the high pressures a compressor reaches, the discharge temperature is found
not from the ideal isentrope but from holding the real entropy $s^{\text{ideal}}
(T,p) + (s - s^{\text{ideal}})$ constant, and the entropy-departure term
computed earlier shifts the discharge temperature by several degrees for a
dense gas such as carbon dioxide near its critical region. The working lesson
is that the ideal isentrope sets the scale of the temperature rise and warns
of auto-ignition, while a compressor duty calculation on a dense gas carries
the entropy departure to place the discharge temperature to the degree the
intercooler sizing needs.

\subsection{Relative humidity after heating a winter air stream}

A heating coil that warms cold outdoor air without adding water drops its
relative humidity, and the drop is a direct reading of the saturation curve
that explains why heated winter interiors feel dry. We carry outdoor air at
$0\,\degC$ and $80\,\%$ relative humidity through a heater to $22\,\degC$ and
find the relative humidity at the room. Heating at constant total pressure
and constant water content holds the vapour partial pressure $p_v$ fixed,
because no water is added or removed, so the humidity ratio does not change.
The inlet vapour pressure is
\[
p_v = \phi_1\,p_{\text{sat}}(0\,\degC) = (0.80)(0.611) = 0.489\,\text{kPa},
\]
using the water saturation pressure of $0.611\,\text{kPa}$ at $0\,\degC$.
After heating to $22\,\degC$, where the saturation pressure has climbed to
$2.645\,\text{kPa}$, the relative humidity is that same vapour pressure
divided by the new, much larger saturation pressure,
\[
\phi_2 = \frac{p_v}{p_{\text{sat}}(22\,\degC)} = \frac{0.489}{2.645} = 0.185,
\]
so the air that left the outdoors at $80\,\%$ relative humidity arrives in
the room at $18\,\%$, drier than a comfortable interior and dry enough to
crack skin and shrink woodwork. The mechanism is the steep rise of the
saturation curve: heating multiplies the saturation pressure by a factor of
$2.645/0.611 = 4.3$ over this span while holding the actual vapour pressure
fixed, so the ratio that is the relative humidity falls by the same factor.
The engineering fix is winter humidification, adding water to the heated
stream to raise $p_v$ back toward a comfortable $30$ to $40\,\%$ relative
humidity, and the humidifier load is the water needed to lift $p_v$ from
$0.489\,\text{kPa}$ to $\phi_{\text{target}}\,p_{\text{sat}}(22\,\degC)$. The
same reading run backward explains summer air conditioning: cooling a stream
without removing water \emph{raises} its relative humidity toward saturation,
which is why an oversized sensible-only cooling coil leaves a space cold and
clammy, the failure the sensible-latent split earlier flagged.

\subsection{The isentropic discharge temperature of a refrigerant compressor}

The refrigeration worked example read the compressor discharge enthalpy off a
superheat table; here we compute the discharge temperature from the property
relations, because a compressor's discharge temperature governs its oil life
and its valve reliability, and an engineer without the exact table needs the
estimate. R-134a leaves the evaporator as saturated vapour at $-20\,\degC$
and is compressed isentropically to $1.0\,\text{MPa}$. Treating the
superheated vapour as ideal with a mean $\gamma \approx 1.10$ over the
compression (R-134a is a heavy polyatomic molecule with many active internal
modes, so $\gamma$ sits close to one) and using the saturation temperature at
$1.0\,\text{MPa}$ of $39.4\,\degC = 312.5\,\si{\kelvin}$ as the state the
vapour is compressed toward, the isentropic temperature rise from the suction
pressure $p_1 = 132.7\,\text{kPa}$ to $p_2 = 1000\,\text{kPa}$ is
\[
T_2 = T_1\left(\frac{p_2}{p_1}\right)^{(\gamma - 1)/\gamma} = 253.15\times
\left(\frac{1000}{132.7}\right)^{0.0909} = 253.15\times (7.536)^{0.0909}.
\]
The pressure-ratio factor is $(7.536)^{0.0909} = 1.205$, so $T_2 = 253.15
\times 1.205 = 305\,\si{\kelvin} = 32\,\degC$. The superheat table returns
about $47\,\degC$ for the true isentropic discharge, so the constant-$\gamma$
ideal estimate undershoots by fifteen degrees, the size of the real-gas and
variable-$\gamma$ error for a heavy refrigerant compressed near its
saturation curve. The undershoot is instructive rather than fatal: it tells
the engineer that a heavy-molecule refrigerant compresses cooler than a
light-gas rule of thumb predicts, which is one reason R-134a and its
successors tolerate high pressure ratios without the discharge temperatures
that would cook the oil in an ammonia or a carbon-dioxide compressor. The
disciplined move is to read the superheat table for the number that sets the
oil-life margin and to keep the ideal estimate as the sanity check that
catches a table look-up in the wrong row.

\subsection{The triple point as a calibration fixed point}

The Gibbs phase rule fixed the triple point of a pure substance at zero
degrees of freedom, and that invariance is what makes the triple point a
metrological fixed point rather than a curiosity. We read why the triple
point of water served as a defining anchor of the temperature scale and what
its zero-freedom character demands of a calibration cell. At the triple point
solid, liquid, and vapour water coexist, so $P = 3$ and $C = 1$ give $F = 1 -
3 + 2 = 0$: neither temperature nor pressure is free, and the three phases
can coexist only at the single point $T = 273.16\,\si{\kelvin}$ ($0.01\,
\degC$) and $p = 611.657\,\text{Pa}$. A sealed glass cell containing pure
water and nothing else, partly frozen so that all three phases are present,
sits at that temperature by thermodynamic necessity, independent of how much
of each phase is present and independent of the laboratory's ambient
conditions, so long as the three phases genuinely coexist. This is why the
triple-point-of-water cell defined the kelvin from $1954$ until the $2019$
redefinition fixed the Boltzmann constant instead: the cell realises one
temperature with no adjustment and no reference to another thermometer,
because the phase rule leaves no knob to turn. The redefinition did not
retire the cell; a triple-point cell remains the most reproducible temperature
realisation near room conditions, good to tens of microkelvin, and it anchors
the practical dissemination of the scale. The working content for an engineer
is the design consequence of $F = 0$: a triple-point cell needs no pressure
control and no precise fill, because the invariance does the regulating,
whereas a boiling-point or a melting-point reference, sitting at $F = 1$ on a
two-phase curve, must control the pressure to hold its temperature. The count
of degrees of freedom is thus not an abstraction but the specification that
separates a self-regulating fixed point from one that needs a barometer.

\subsection{A vignette: the espresso machine and flash boiling}

A steam wand and a pump-driven espresso group are two readings of the
saturation curve inside one appliance, and the machine repays a close look
because its whole behaviour turns on where the water sits relative to its
boiling point. The boiler holds water under pressure so that it can store
heat above the atmospheric boiling point without flashing: a domestic machine
runs its steam boiler near $130\,\degC$ at a gauge pressure of about
$1.7\,\text{bar}$, read straight off the saturation curve as the pressure at
which water boils at $130\,\degC$. When the steam valve opens to froth milk,
the pressurised vapour leaves the boiler, expands to atmospheric, and delivers
its latent heat to the milk as it recondenses on the cold surface, the same
condensing-steam heat transfer the autoclave used, here dumping roughly
$2200\,\text{kJ}/\text{kg}$ into the pitcher and warming and foaming the milk
in seconds. The brew group works the other side of the curve. The pump
drives water at about $9\,\text{bar}$ through the coffee bed at a temperature
near $93\,\degC$, deliberately below the atmospheric boiling point so the
water stays liquid through the puck and extracts rather than steams; the high
pressure is for flow through the fine grind, not for boiling. The failure mode
the machine designs against is flash boiling at the group. If the brew water
is too hot, or the pressure drops too fast as the shot ends, the water crosses
its saturation curve on the way out of the puck and flashes to vapour,
spitting and channelling and pulling a sour, under-extracted shot. The cure is
the same margin a pump keeps from cavitation: hold the brew water a few
degrees below its saturation temperature at the group pressure, and bleed the
pressure down gently at the end of the shot rather than releasing it at once.
The machine is the chapter in a kitchen: one appliance that stores heat above
the boiling point by sealing the vapour space, delivers it by condensing steam,
and brews by keeping the water a controlled margin on the liquid side of the
saturation curve, failing exactly when it lets the water cross to the vapour
side at the wrong moment.

\subsection{A vignette: freeze-drying and the sublimation leg}

Freeze-drying removes water from a heat-sensitive product by routing it around
the liquid phase entirely, and the route is a path on the $p$-$T$ diagram that
stays below the triple point. The product, a vaccine, a culture, an instant
coffee, is first frozen solid, then placed in a chamber pumped to a pressure
below water's triple-point pressure of $611\,\text{Pa}$, typically to tens of
pascals. At that pressure water has no liquid field on its phase diagram: the
saturation curve that separates liquid from vapour exists only above the
triple point, so below it the solid passes directly to vapour along the
sublimation curve, and the ice in the frozen product sublimes without ever
melting. The sublimation draws its latent heat, the sum of the fusion and
vapourisation latent heats, about $2838\,\text{kJ}/\text{kg}$ for water,
from the product, which a gentle shelf heater supplies while the product
stays cold because the subliming ice pins its temperature the way boiling
pins an open pot. The vapour is pumped away and refrozen on a cold condenser
coil held near $-50\,\degC$, whose saturation pressure is low enough to keep
driving the sublimation. The whole cycle is a deliberate tour of the phase
diagram that avoids the liquid: freeze to enter the solid field, pump below
the triple point to reach the sublimation curve, sublime the ice to vapour at
low temperature, and recapture the vapour as ice on a colder surface. The
engineering payoff is that the product never sees the liquid water that would
carry dissolved solutes into damaging concentrations, never sees the
high temperature that would denature a protein, and dries to a porous solid
that rehydrates fast. The design constraints all read off the diagram: the
chamber pressure must stay below $611\,\text{Pa}$ or the ice melts and the
batch is lost, the shelf heat must match the sublimation latent heat without
warming the product above the temperature at which the partly-dried matrix
collapses, and the condenser must run cold enough that its saturation pressure
stays below the subliming surface's. Freeze-drying is sublimation engineered,
the one leg of the phase diagram the rest of the chapter passed over.

\subsection{The latent heat that a boiler carries and the vanishing at the
critical point}

A boiler warms its feedwater, boils it, and superheats the vapour, and the
three legs draw very different amounts of heat, so an engineer sizing the
duty needs to know how the latent leg compares with the sensible legs and how
both change with the operating pressure. We size the three legs for water at
$8\,\text{MPa}$, the pressure of the Rankine case study, and read the latent
heat off its temperature dependence. The saturation temperature at
$8\,\text{MPa}$ is $295\,\degC$, so the feedwater warms from the condenser
return near $46\,\degC$ across a $249\,\si{\kelvin}$ span. The
subcooled-liquid sensible heat is
\[
q_{\text{sens,liq}} \approx c\,(T_{\text{sat}} - T_{\text{in}}) = (4.6)(295 -
46) = 1145\,\text{kJ}/\text{kg},
\]
using a mean liquid heat capacity of $c \approx 4.6\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$ that runs above the room-temperature $4.18$ because the liquid
warms toward the critical region. The latent leg at $8\,\text{MPa}$ carries
$h_{fg} = 1442\,\text{kJ}/\text{kg}$, read off the falling curve of
figure~\ref{fig:vol04ch02:latent-heat-temperature} at the $295\,\degC$
saturation point, and the superheat leg from $295\,\degC$ to $500\,\degC$
carries about $q_{\text{super}} \approx c_p\,(500 - 295) = (2.6)(205) =
533\,\text{kJ}/\text{kg}$ with a superheated-vapour heat capacity near
$2.6\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$. The three sum to $1145 + 1442 +
533 = 3120\,\text{kJ}/\text{kg}$, within a few percent of the table's
$q_{\text{in}} = 3199\,\text{kJ}/\text{kg}$, the difference being the
temperature averaging of the heat capacities. The reading that matters is the
shrinking of the latent leg with pressure: at atmospheric pressure the latent
heat is $2257\,\text{kJ}/\text{kg}$ and dwarfs the sensible legs, while at
$8\,\text{MPa}$ it has fallen to $1442$ and is now comparable to the liquid
warming, and at the critical pressure of $22.06\,\text{MPa}$ it vanishes
entirely, as the curve of figure~\ref{fig:vol04ch02:latent-heat-temperature}
shows falling to zero at $T_c$. A supercritical boiler exploits exactly this
vanishing: above the critical pressure there is no latent plateau at all, the
fluid warms continuously from liquid-like to gas-like without a phase change,
and the boiler duty is one smooth sensible ramp. The latent heat that pins a
subcritical boiler at its saturation temperature is the same quantity whose
disappearance lets a supercritical unit dispense with the two-phase leg
altogether.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-latent-heat-temperature.tex}

\subsection{The boiling profile of a heat exchanger and its pinch}

An engineer who must transfer heat into a boiling stream reads the load not as
a single number but as a profile against temperature, because the boiling leg
sits at one fixed temperature and constrains how close the heating stream can
approach. We trace the temperature-against-heat profile of the same
$8\,\text{MPa}$ boiler and locate the pinch it imposes. Following the three
legs of the previous example, the water enters at $46\,\degC$ and warms as
subcooled liquid to the saturation temperature of $295\,\degC$ over the first
$q_{\text{sat}} = 1145\,\text{kJ}/\text{kg}$ of heat (the subcooled-liquid
leg, above a room-temperature $c\,\Delta T$ estimate because the liquid heat
capacity climbs toward the critical region), then boils across a flat plateau
at $295\,\degC$ absorbing the
$h_{fg} = 1442\,\text{kJ}/\text{kg}$ the table gives at this pressure, then
superheats to $500\,\degC$ over the remaining heat, the profile drawn in
figure~\ref{fig:vol04ch02:boiler-tq-profile}. The flat boiling plateau is the
largest single block of heat and the reason a boiler spends most of its length
at one temperature. The plateau also sets the pinch point of a counter-flow
heat exchanger. A hot flue gas or a hot process stream heating this water must
stay above the working-fluid temperature at every point along the exchanger, so
the tightest approach occurs where the cold stream is flat and the hot stream is
still falling: at the start of boiling, the pinch. If the hot stream is a flue
gas entering at $600\,\degC$ and leaving at $150\,\degC$, its temperature
crosses toward the flat $295\,\degC$ plateau, and the minimum approach there,
often specified at $10$ to $20\,\si{\kelvin}$, fixes how much of the flue gas's
heat the boiler can recover. The working content is that a two-phase load
cannot be sized on total duty alone; the flat plateau of
figure~\ref{fig:vol04ch02:boiler-tq-profile} is a geometric constraint the hot
stream must clear at the pinch, and an exchanger sized on average temperature
difference will pinch and stall at the boiling leg.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-boiler-tq-profile.tex}

\subsection{A saturated-steam accumulator sized from the flash fraction}

A steam accumulator stores energy as hot pressurised water and releases it as
steam when the pressure is dropped, and the store an engineer can draw from it
is the flash the pressure drop produces, the same isenthalpic flash the
refrigeration throttle showed, run on water for a different purpose. We size
the useful steam a Ruths accumulator delivers. The vessel holds saturated
water at a charged pressure of $1.5\,\text{MPa}$, where the saturation
temperature is $198\,\degC$ and the saturated-liquid enthalpy is $h_f(1.5\,
\text{MPa}) = 845\,\text{kJ}/\text{kg}$. When a process draws steam and the
pressure falls to the discharge pressure of $0.4\,\text{MPa}$ (saturation
temperature $144\,\degC$), the water finds itself above its new saturation
temperature and flashes: each kilogram of stored water holds its enthalpy
$845\,\text{kJ}/\text{kg}$ constant while dropping to the lower pressure,
where $h_f(0.4\,\text{MPa}) = 605\,\text{kJ}/\text{kg}$ and $h_g(0.4\,
\text{MPa}) = 2739\,\text{kJ}/\text{kg}$. The flash fraction is the quality
the water reaches at the discharge pressure,
\[
x_{\text{flash}} = \frac{h_f(1.5\,\text{MPa}) - h_f(0.4\,\text{MPa})}{h_g(0.4
\,\text{MPa}) - h_f(0.4\,\text{MPa})} = \frac{845 - 605}{2739 - 605} =
\frac{240}{2134} = 0.112,
\]
so $11.2\,\%$ of each kilogram of stored water flashes to steam on the pressure
swing. An accumulator charged with $20{,}000\,\text{kg}$ of water therefore
delivers $0.112\times 20{,}000 = 2240\,\text{kg}$ of steam over the swing from
$1.5$ to $0.4\,\text{MPa}$, buffering a process that draws steam faster than the
boiler can raise it. The working lesson is that the store is not the total water
mass but the flash fraction that mass yields between the charge and discharge
pressures, which is why an accumulator is sized by the pressure swing it can
tolerate: a wider swing flashes a larger fraction and stores more steam per
kilogram of water, while a process that can tolerate only a narrow pressure
band needs a proportionally larger water inventory for the same steam delivery.

\subsection{The frost point below freezing and a cold-surface condensation
check}

The dew-point construction found the temperature at which vapour condenses to
liquid, but below freezing the vapour deposits directly as frost, and the
relevant curve is the sublimation curve rather than the vapourisation curve. We
find the frost point of cold outdoor air and check whether it will rime a
heat-exchanger surface. Outdoor air sits at $-5\,\degC$ and $70\,\%$ relative
humidity, where relative humidity is referenced to the saturation pressure over
ice below freezing. The saturation pressure of water vapour over ice at
$-5\,\degC$ is $0.402\,\text{kPa}$, so the actual vapour partial pressure is
\[
p_v = \phi\,p_{\text{sat,ice}}(-5\,\degC) = (0.70)(0.402) = 0.281\,\text{kPa}.
\]
The frost point is the temperature at which the ice-saturation pressure falls to
this vapour pressure. Reading the sublimation curve, the saturation pressure
over ice equals $0.281\,\text{kPa}$ at about $-9\,\degC$, so the frost point is
$-9\,\degC$. Any surface colder than $-9\,\degC$ in this air stream, the fin of
an air-source heat pump's outdoor coil running below the frost point on a cold
damp day, deposits frost directly from the vapour without a liquid stage, and
the frost blanket insulates the coil and throttles the airflow until a defrost
cycle clears it. The construction is the sublimation-curve twin of the
dew-point calculation: above freezing the vapour meets the vapourisation curve
and sweats as liquid, below freezing it meets the sublimation curve and rimes as
frost, and the design rule for a cold coil is to keep its surface above the
frost point of the incoming air or to schedule a defrost that melts the
accumulated rime. The reason an air-source heat pump loses capacity in cold damp
weather is exactly this frosting: the coil must run below the air's frost point
to pull heat from it, so it frosts, and the periodic defrost that clears the
frost is itself a debit against the heat delivered.

\subsection{The relative-volatility limit on a single equilibrium stage}

The binary flash and the Raoult bubble-point both enriched a vapour above its
feed, and the enrichment a single stage can reach is bounded by the relative
volatility, the number that decides whether one stage suffices or a column is
needed. We read that bound for an ethanol-water stream and see why the mixture
defeats it. The relative volatility of the light over the heavy component is
$\alpha = K_1/K_2 = (y_1/x_1)/(y_2/x_2)$, and for a binary it rearranges to the
single-stage enrichment relation
\[
y_1 = \frac{\alpha\,x_1}{1 + (\alpha - 1)\,x_1},
\]
which gives the vapour composition a single equilibrium stage produces from a
liquid of composition $x_1$. For a benzene-toluene liquid with $\alpha = 2.48$
and $x_1 = 0.40$ the stage yields $y_1 = (2.48)(0.40)/[1 + (1.48)(0.40)] =
0.992/1.592 = 0.623$, the same $62\,\%$ the Raoult bubble-point example
returned, confirming the two routes agree. The bound the relation carries is
that as $\alpha \to 1$ the enrichment $y_1 \to x_1$ and the stage does nothing,
which is why a low relative volatility forces many stages. Ethanol and water
carry the sharper lesson: their relative volatility falls as the ethanol
concentration rises, and at $x_1 = 0.894$ mole fraction ethanol the relative
volatility reaches exactly one, $\alpha = 1$, so $y_1 = x_1$ and no
distillation can enrich the stream further. This is the ethanol-water
\engterm{azeotrope}, the composition at which the vapour and liquid share a
composition and ordinary distillation stalls at about $95\,\%$ ethanol by
volume. The working content is that the single-stage enrichment relation, run
with a composition-dependent $\alpha$, both sizes the enrichment a real stage
delivers and locates the azeotrope where the enrichment vanishes, and it is why
producing anhydrous ethanol needs a step beyond simple distillation, a
molecular sieve or an entrainer that breaks the azeotrope the relative
volatility of one imposes.

\subsection{A vignette: the multi-stage compressor and intercooling}

A gas compressed in one stroke heats up, and the heat is a debit the designer
pays twice: once as compression work that the temperature rise inflates, and
once as a discharge temperature that threatens the machine. Splitting the
compression into stages and cooling the gas between them recovers both, and the
device that does it is the multi-stage intercooled compressor that raises air
for a plant, a pipeline, or a bottling line. The physics is the area under the
compression path on the $p$-$v$ plane. Compressing a gas from a suction
pressure $p_1$ to a discharge pressure $p_3$ in a single adiabatic stroke
follows the steep adiabat $p v^{\gamma} = \text{const}$, and the work per unit
mass is the area between that path and the pressure axis, which for an ideal gas
is
\[
w_{\text{single}} = \frac{\gamma}{\gamma - 1} R_s T_1\left[\left(\frac{p_3}{p_1}
\right)^{(\gamma - 1)/\gamma} - 1\right].
\]
For air drawn at $T_1 = 300\,\si{\kelvin}$ and compressed through an overall
pressure ratio of $9$ with $\gamma = 1.4$ and $R_s = 287\,\text{J}/(\text{kg}
\cdot\si{\kelvin})$, the pressure-ratio factor is $9^{0.2857} = 1.873$, so the
single-stage work is $w_{\text{single}} = (3.5)(287)(300)(0.873) = 263\,
\text{kJ}/\text{kg}$, and the discharge temperature is $T_3 = (300)(1.873) =
562\,\si{\kelvin} = 289\,\degC$, hot enough to carbonise the lubricating oil
and to risk a compression-ignition fire in an oil-fouled machine.

Splitting the compression at an intermediate pressure and cooling the gas back
to its inlet temperature between the two stages cuts both. The work is smallest
when the two stages share the pressure ratio equally, $p_2/p_1 = p_3/p_2 =
\sqrt{p_3/p_1} = 3$, a result that falls out of minimising the sum of the two
stage works over the split pressure. Each stage then does the work of a
pressure ratio of $3$, starting from the same inlet temperature because the
intercooler restores it, so the total is
\[
w_{\text{two-stage}} = 2\times\frac{\gamma}{\gamma - 1} R_s T_1\left[3^{(\gamma
- 1)/\gamma} - 1\right] = 2(3.5)(287)(300)(3^{0.2857} - 1) =
2(3.5)(287)(300)(0.369) = 222\,\text{kJ}/\text{kg},
\]
a $16\,\%$ saving on the compression work, and each stage discharges at only
$T = (300)(1.369) = 411\,\si{\kelvin} = 138\,\degC$, well below the oil-damage
threshold. The path on the $p$-$v$ plane, drawn in
figure~\ref{fig:vol04ch02:two-stage-intercooling}, shows why: the single
adiabat climbs steeply and encloses a large area, while the two-stage path
compresses, steps back to a smaller volume along the constant-pressure
intercooling line, and compresses again, hugging closer to the isotherm that is
the lowest-work path a perfectly cooled compressor could follow. The area
between the single adiabat and the two-stage path is the work the intercooler
saves. Adding a third and a fourth stage drives the path still closer to the
isotherm, each stage cutting a smaller further slice of work at the cost of
another heat exchanger, which is why a plant air compressor is typically two or
three stages and a high-pressure breathing-air or a natural-gas pipeline
compressor may run four or more. The equation of state sets the whole trade: the
adiabat's steepness is $\gamma$, the isotherm is the ideal-gas $pv =
\text{const}$, and the intercooler's job is to keep dragging the real path down
toward that isotherm one stage at a time.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-two-stage-intercooling.tex}

\begin{mastery}
The van der Waals second virial coefficient of the Boyle-temperature example
is the first term of a systematic connection between the equation of state and
the intermolecular potential that statistical mechanics makes exact. The
second virial coefficient is
\[
B(T) = -2\pi N_A \int_0^\infty \left[e^{-u(r)/k_B T} - 1\right] r^2\,dr,
\]
with $u(r)$ the pair potential, $N_A$ Avogadro's number, and $k_B$ the
Boltzmann constant. The integrand weighs the Boltzmann factor of the pair
interaction against the no-interaction reference of unity, so a region of
attraction ($u < 0$) contributes positively to the integral and pulls $B$
negative, while the hard-core repulsion ($u \to \infty$ at small $r$) makes
the integrand $-1$ over the excluded sphere and pushes $B$ positive. For a
hard-sphere potential of diameter $\sigma$ the integral is elementary and
gives $B = (2\pi/3) N_A \sigma^3$, the excluded volume, temperature
independent, which is the statistical-mechanical meaning of the van der Waals
$b$. Adding a weak attractive tail contributes a negative $-a/(RT)$ term, and
the two together reproduce $B(T) = b - a/(RT)$, so the van der Waals
coefficients are not fitted fudge factors but the leading hard-core and
attractive moments of the real pair potential. The higher virial coefficients
$C(T)$, $D(T)$ carry three-body and four-body cluster integrals, and the
series is the rigorous low-density expansion whose truncation error an
engineer controls by staying below moderate density. The lesson is that the
cubic equations of state trade this rigour for range: the virial equation is
exact term by term but converges only at low density, while a fitted cubic
extrapolates through the critical region at the cost of the clean
potential-to-coefficient correspondence the virial series keeps.
\end{mastery}

\begin{mastery}
The lever rule and the quality read a two-phase state's phase split off a
single measured property, but the same construction extends to the entropy
and carries the second-law content of a phase change. Along an isotherm-isobar
inside the dome, the entropy of vapourisation is $s_{fg} = s_g - s_f =
h_{fg}/T$, the reversible latent heat divided by the absolute temperature, so
the width of the dome on a $T$-$s$ diagram at a given temperature is exactly
$h_{fg}/T$. This is why the vapour dome narrows toward the critical point: as
$T \to T_c$ the latent heat $h_{fg} \to 0$ faster than $T$ falls, so $s_{fg}
\to 0$ and the two limbs of the dome meet. The entropy reading also fixes the
quality from a $T$-$s$ state directly, $x = (s - s_f)/s_{fg}$, the same lever
rule applied on the entropy axis, which is the form the turbine and Rankine
examples used to find the wet-exit quality from the isentropic entropy. A
deeper consequence is the shape of the saturated-vapour limb. For most fluids
the limb leans to the left with increasing entropy at lower temperature, so an
isentropic expansion from dry saturated vapour drops into the two-phase
region and wets, the behaviour that erodes a steam turbine's last stages. A
small class of heavy, complex-molecule fluids (many of the working fluids of
organic Rankine cycles) have a saturated-vapour limb that leans the other way,
so an isentropic expansion from saturation moves \emph{into} the superheated
region and stays dry. These \engterm{retrograde} or dry fluids need no
superheat to protect the expander from droplets, which is one reason a
waste-heat organic Rankine unit can run a simple saturated cycle where a steam
plant must superheat and reheat. The sign of the saturated-vapour limb's slope,
read off the $T$-$s$ diagram, is thus a working fluid-selection criterion, and
it traces back through $s_{fg} = h_{fg}/T$ to how the latent heat and the
heat capacity of the fluid trade off along its saturation curve.
\end{mastery}

\begin{mastery}
The reduced-pressure swing of the compressibility factor that the chart
records is also where the simple equations of state earn their keep. A
single van der Waals constant pair reproduces the qualitative shape of the
single van der Waals constant pair reproduces the qualitative shape of the
$Z$ curve but misses the depth of the critical dip by a wide margin: the
van der Waals critical compressibility is $Z_c = p_c v_c/(R T_c) = 3/8 =
0.375$ for every substance, while measured values cluster near $0.27$ to
$0.29$ (see \texttt{data/critical\_constants.csv}). The cubic equations of
state used in industry (Redlich-Kwong, Soave-Redlich-Kwong,
Peng-Robinson) add a temperature-dependent attraction term and an
acentric-factor correction precisely to bend $Z_c$ down toward the
measured value and to fit vapour-liquid equilibrium. The lesson for the
working engineer is to read the van der Waals equation as a teaching
model that gets the phenomenology right and the numbers wrong, and to
reach for a fitted cubic or a reference equation of state (REFPROP,
CoolProp) when the numbers must be trusted to a few percent.
\end{mastery}

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-zchart-corresponding-states.tex}

\begin{mastery}
The cubic equations of state that industry trusts share one move: they
keep the van der Waals two-parameter skeleton and bend it with a
temperature-dependent attraction and a third substance descriptor, the
\engterm{acentric factor} $\omega$. Pitzer defined $\omega$ from the
saturation curve itself,
\[
\omega = -1 - \log_{10}\!\left(\frac{p_{\text{sat}}}{p_c}\right)_{T_r =
0.7},
\]
the displacement of a substance's reduced saturation pressure at
$T_r = 0.7$ below the value $0.1$ that the simple-fluid argon-krypton-xenon
family takes. Spherical non-polar molecules have $\omega \approx 0$
(argon $0.000$, methane $0.011$); elongated or mildly polar molecules
have larger $\omega$ (propane $0.152$, water $0.344$, R-134a $0.327$).
The Peng-Robinson equation,
\[
p = \frac{RT}{v - b} - \frac{a\,\alpha(T)}{v(v + b) + b(v - b)},
\]
folds $\omega$ into the temperature function $\alpha(T) = [1 +
\kappa(1 - \sqrt{T_r})]^2$ with $\kappa = 0.37464 + 1.54226\,\omega -
0.26992\,\omega^2$. The reward is a predicted critical compressibility of
$Z_c = 0.307$ against the van der Waals $0.375$, much closer to the
measured cluster near $0.27$ to $0.29$ that
figure~\ref{fig:vol04ch02:zchart-corresponding-states} marks, plus a
vapour-liquid equilibrium that tracks hydrocarbon mixtures to within a
few percent. The principle of corresponding states is the reason a single
chart, indexed by $(p_r, T_r)$ and refined by $\omega$, serves for most
non-polar fluids: two substances at the same reduced state with the same
acentric factor occupy nearly the same reduced volume. The principle is
an experimental regularity, exact only for substances sharing an
intermolecular potential, which is why it works well for the hydrocarbon
series and poorly for hydrogen-bonding fluids such as water and ammonia,
whose departures the acentric factor only partly absorbs.
\end{mastery}

\begin{mastery}
The throttling example held enthalpy constant and watched a liquid flash,
but a throttle acting on a single-phase gas does something subtler: it
changes the temperature, and the sign of that change is the
\engterm{Joule-Thomson coefficient}
\[
\mu_{JT} = \left(\frac{\partial T}{\partial p}\right)_h
= \frac{1}{c_p}\left[T\left(\frac{\partial v}{\partial T}\right)_p -
v\right].
\]
For an ideal gas $v = R_s T/p$ makes the bracket vanish identically, so
an ideal gas neither warms nor cools on throttling. A real gas has a
non-zero coefficient whose sign flips along the \engterm{inversion
curve}, the locus where $\mu_{JT} = 0$. Inside the inversion curve
$\mu_{JT} > 0$ and a pressure drop cools the gas; outside it $\mu_{JT} <
0$ and a pressure drop warms the gas. The cooling region is the working
basis of gas liquefaction: the Linde cycle throttles a high-pressure gas
that sits inside its inversion curve, takes the small temperature drop,
recuperates it against the incoming stream, and repeats until the gas
condenses. The hazard hides in the sign flip. Hydrogen and helium have
inversion temperatures far below room temperature ($202\,\si{\kelvin}$
for hydrogen, $40\,\si{\kelvin}$ for helium), so at ambient conditions
they sit outside the inversion curve and a throttling leak \emph{warms}
them, which is why a high-pressure hydrogen release can self-ignite at a
pinhole while a throttled carbon-dioxide or nitrogen leak (both with
room-temperature $\mu_{JT} > 0$) cools and can frost the fitting. The van
der Waals equation predicts the inversion curve qualitatively, giving a
maximum inversion temperature of $T_{\text{inv}}^{\max} = 2a/(R b) =
(27/4)\,T_c$ in the low-pressure limit, which places nitrogen's at about
$850\,\si{\kelvin}$ and hydrogen's at about $220\,\si{\kelvin}$, both in
the right neighbourhood of the measured values. The lesson for a designer
sizing a let-down station is to know which side of the inversion curve
the gas sits on before assuming a throttle cools it.
\end{mastery}

\begin{mastery}
The transcritical case study rejected heat from a supercritical fluid that
never condensed, which raises the question of where, above the critical
point, the fluid stops behaving like a liquid and starts behaving like a
gas. There is no sharp line, but there is a soft one, the \engterm{Widom
line}, the locus of maxima of the isobaric heat capacity $c_p$ as a function
of temperature at each supercritical pressure. Below the Widom line the
supercritical fluid is dense and liquid-like, with high density and low
compressibility; above it the fluid is dilute and gas-like. The two
behaviours meet at the critical point and the Widom line emanates from it as
the smooth continuation of the saturation curve into the single-phase
region. The crossing carries no latent heat, because there is no phase
transition, but $c_p$ peaks sharply near the line, so a supercritical heat
exchanger that cools its fluid across the Widom line passes through a region
of large effective heat capacity that acts like a smeared-out latent heat.
This is exactly what a carbon-dioxide gas cooler does: cooling the
high-side fluid at, say, $9\,\text{MPa}$ from $80\,\degC$ to $35\,\degC$
carries it across the Widom line near $40\,\degC$, where the steep $c_p$
peak lets the gas cooler reject a large enthalpy over a small temperature
span. The sharp temperature dependence of the supercritical properties is
the reason the gas-cooler outlet temperature matters so much in the
transcritical optimisation: a few degrees on the outlet temperature, near
the Widom line, moves the exit enthalpy a great deal, and the
near-linear optimal-pressure rule of thumb is the engineer's way of tracking
the peak without computing the full property surface at every operating
point. The lesson for the designer is that the critical point is not a wall
but a fading boundary, and a supercritical heat exchanger is engineered
around the Widom-line $c_p$ peak the way a subcritical one is engineered
around the latent heat. Figure~\ref{fig:vol04ch02:widom-cp-peak} draws the
peak at three supercritical pressures and marks the Widom line as the locus
of the maxima.
\end{mastery}

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-widom-cp-peak.tex}

\begin{mastery}
The cubic equations of state guess a functional form and fit it; the virial
equation makes no guess, expanding the compressibility factor as a power
series in density whose coefficients are tied to the intermolecular
potential itself,
\[
Z = \frac{pv}{RT} = 1 + \frac{B(T)}{v} + \frac{C(T)}{v^2} + \cdots,
\]
with $B(T)$ the second virial coefficient, $C(T)$ the third, and so on.
Statistical mechanics gives the coefficients their meaning: $B(T)$ measures
the net effect of interactions between pairs of molecules, $C(T)$ between
triples, and the series is the systematic low-density expansion of the gas.
The second coefficient carries the competition between repulsion and
attraction directly. At low temperature the slow molecules linger in one
another's attractive wells, the gas is easier to compress than ideal, and
$B(T) < 0$; at high temperature the fast molecules feel mostly the
hard-core repulsion, the gas is harder to compress than ideal, and $B(T) >
0$. The temperature where the two balance and $B(T) = 0$ is the
\engterm{Boyle temperature} $T_B$, at which a real gas obeys $pv = RT$ to
first order in density over a wide pressure range, behaving ideally not
because it is dilute but because its pairwise interactions cancel. For the
van der Waals equation the second virial coefficient follows from expanding
$p = RT/(v-b) - a/v^2$ in powers of $1/v$: writing $1/(v-b) = (1/v)(1 +
b/v + \cdots)$ gives $Z = 1 + (b - a/RT)/v + \cdots$, so $B(T) = b - a/(RT)$
and the Boyle temperature is $T_B = a/(Rb)$. With the critical-point
relations $a = 27R^2T_c^2/(64p_c)$ and $b = RT_c/(8p_c)$ this is $T_B =
(27/8)\,T_c = 3.4\,T_c$, which places nitrogen's Boyle temperature near
$430\,\si{\kelvin}$ against a measured $327\,\si{\kelvin}$, the right
neighbourhood with the usual van der Waals overshoot. The working value of
the truncated virial equation is at low to moderate density, where two
coefficients suffice and the series converges quickly; near and above the
critical density the expansion fails and an engineer reaches for a cubic or
a reference equation of state, which is why the virial form is the tool of
thermophysical reference work and the cubics are the tools of process
design.
\end{mastery}

\begin{mastery}
The spinodal marks the absolute limit of a metastable phase, but a real
liquid rarely reaches it, because a bubble usually nucleates first,
somewhere in the gap between the saturation line and the spinodal that
figure~\ref{fig:vol04ch02:spinodal-superheat} draws. The obstacle is
surface energy. Forming a vapour bubble of radius $r$ inside a superheated
liquid pays a volume reward proportional to $r^3$ (the pressure difference
times the bubble volume) but a surface cost proportional to $r^2$ (the
surface tension times the bubble area), so the free energy of a nascent
bubble rises before it falls, and a bubble smaller than the critical radius
$r^* = 2\sigma/\Delta p$ collapses rather than grows. The height of that
barrier sets the nucleation rate, which climbs by many orders of magnitude
over a narrow temperature band as the superheat deepens and $\Delta p$
grows, so a superheated liquid does nothing for a long time and then boils
suddenly, the behaviour a bumping flask shows and a smooth
microwave-heated cup of water shows more dangerously. Two regimes divide the
metastable gap. \engterm{Heterogeneous nucleation} on a wall crevice, a
dust mote, or a dissolved gas pocket needs far less superheat, because the
foreign surface supplies part of the bubble wall and cuts the barrier; this
is ordinary boiling, and it keeps a kettle near its saturation temperature.
\engterm{Homogeneous nucleation}, in a liquid clean of nucleation sites,
holds out almost to the spinodal, to the limit of superheat the previous
worked example computed, and it is the regime a BLEVE enters when a vessel
depressurises faster than heterogeneous sites can respond: the whole liquid
mass, thrown deep into the metastable region at once, nucleates
homogeneously and everywhere, and the flash is explosive rather than
orderly. The design reading is that the saturation line is where boiling may
begin, the spinodal is where it must, and the distance between them, governed
by surface tension and the population of nucleation sites, is the margin a
process either exploits (a smooth evaporator tube that suppresses premature
boiling) or fears (a pressure vessel whose sudden failure denies the liquid
its nucleation sites and drives it to the spinodal).
\end{mastery}

\begin{mastery}
The stability archetype that the two-phase region illustrates has a general
statement in the convexity of the free energy, and reading it out shows why the
spinodal sits where it does and why the same criterion forbids two other
absurdities. Thermodynamic stability requires that a system resist any
fluctuation by responding in the direction that damps it, and the second law
turns that requirement into curvature conditions on the free energy. For a
simple compressible substance the internal energy $u(s, v)$ must be a convex
function of its natural variables, which forces two inequalities. The first is
mechanical stability, $(\partial p/\partial v)_T < 0$: a compression must raise
the pressure, so that a fluid squeezed by a fluctuation pushes back rather than
collapsing. The second is thermal stability, $c_v > 0$: adding heat at constant
volume must raise the temperature, so that a warm fluctuation does not draw yet
more heat and run away. The isothermal compressibility $\kappa_T =
-(1/v)(\partial v/\partial p)_T$ is therefore positive in any stable state, and
its divergence marks the stability limit. The van der Waals loop violates the
first inequality along its rising branch, where $(\partial p/\partial v)_T >
0$, and the spinodal is exactly the locus $(\partial p/\partial v)_T = 0$ where
$\kappa_T$ diverges and the homogeneous fluid loses mechanical stability; the
worked spinodal example located it as the far edge of the metastable region for
this reason. A second consequence is a bound on the heat capacities. The
general relation $c_p - c_v = T v \beta^2/\kappa_T$, with $\beta$ the thermal
expansion coefficient, is a ratio of two positive quantities scaled by
$\beta^2$, so $c_p \ge c_v$ always, with equality only where $\beta = 0$ (liquid
water at its $4\,\degC$ density maximum is the textbook case). The same relation
diverges as $\kappa_T \to 0$ approaching the spinodal from the stable side,
which is the thermodynamic root of the $c_p$ peak the Widom-line box described:
near the critical point $\kappa_T$ grows large and $c_p$ with it. The lesson is
that the stability of matter, the sign of a pump's response to cavitation, the
location of the spinodal, and the peaks of the heat capacities are one set of
inequalities read in different places, all of them the convexity of the free
energy that the second law demands.
\end{mastery}

\begin{estimation}
\textbf{Estimate the size (volume) and pressure of a propane tank that
holds enough fuel to heat a $200\,\text{m}^2$ Canadian house for one
winter month at average outdoor $-10\,\degC$.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The heat-loss rate of a moderately-insulated $200\,\text{m}^2$ house
at $\Delta T = 30\,\si{\kelvin}$ above ambient is roughly
$8\,\text{kW}$ (one $1\,\text{kW}$ of heat-loss per $25\,\text{m}^2$
of floor at this temperature difference). Over a month of $30 \times
24 = 720\,\text{hours}$, the total heat is $8 \times 720 = 5760\,
\text{kWh} \approx 20\,\text{GJ}$.

Propane has a lower heating value of $46\,\text{MJ}/\text{kg}$, so
the required mass is $20\,\text{GJ}/(46\,\text{MJ}/\text{kg}) \approx
435\,\text{kg}$. The density of liquid propane at storage conditions
($\sim 0.51\,\text{kg}/\text{L}$) gives a volume of $850\,\text{L}$;
a $1000\,\text{L}$ residential tank, the standard size, would suffice
with margin.

The tank's working pressure is the saturation pressure of propane at
the ambient temperature. At $-10\,\degC$, propane's saturation
pressure is $\sim 280\,\text{kPa}$ (gauge); at $30\,\degC$ on a hot
summer day, $\sim 1000\,\text{kPa}$ (gauge). Residential propane
tanks are rated to about $2.5\,\text{MPa}$ to accommodate the highest
foreseeable storage temperature.

The published storage figure for a Canadian house in this climate
zone is $4{,}000$ to $7{,}000\,\text{kWh}/\text{month}$ across mild
to harsh winter months \cite{gen:mackay2009}; our estimate at $5760\,
\text{kWh}$ falls in the middle of the range.

The postmortem: the working calculation gets the tank size right
within a factor of two. The largest uncertainty is the house's actual
heat-loss rate, which depends on insulation, glazing, occupancy, and
infiltration. The propane heating value and density are well-fixed.
A house with substantially better insulation (passive house, factor
of $4$ reduction in heat-loss rate) would need a quarter of the
propane.
\end{estimation}

\section{An end-to-end case study: sizing a residential propane
tank}\pathtag{standard}

The chapter's apparatus pays off when an engineer must size a real vessel
from a duty and clear it for service. We take a residential propane
(LPG) tank for a house heated by a propane furnace, and carry it from the
heating duty through the liquid charge, the fill limit, the pressure the
tank holds across the weather band, the relief-valve sizing for a fire,
and the BLEVE margin, ending in a specification an engineer would sign.
The propane tank is a good vehicle because every move of the chapter
appears in it once: the ideal-gas law for the vapour, the saturation
curve for the working pressure, the liquid density and the
incompressibility approximation for the fill limit, the latent heat for
the relief duty, and the two-phase flash that the failure section turns
dangerous.

\subsection{Duty and the liquid charge}

The house is a moderately insulated $200\,\text{m}^2$ dwelling in a
climate with a design heating load we take at $12\,\text{kW}$ on the
coldest day and an average winter draw of $6\,\text{kW}$ across the
heating season. We size the tank for a one-month refill interval at the
average draw, with the coldest-day load checked against the vapour
off-take rate later. The monthly heat demand is
\[
Q_{\text{month}} = (6\,\text{kW})(30\times 24\,\text{h}) = 4320\,
\text{kWh} = 1.555\times 10^{10}\,\text{J} = 15.6\,\text{GJ}.
\]
A propane furnace at a seasonal efficiency of $0.90$ must burn fuel of
heating value $H = 46\,\text{MJ}/\text{kg}$, so the propane mass per
month is
\[
m = \frac{Q_{\text{month}}}{\eta\, H} = \frac{1.555\times 10^{10}}
{(0.90)(46\times 10^6)} = 376\,\text{kg}.
\]
The liquid density of propane at the $15\,\degC$ fill temperature is
$\rho_f = 507\,\text{kg}/\text{m}^3$, so the liquid volume of the charge
is
\[
V_{\text{liq}} = \frac{m}{\rho_f} = \frac{376}{507} = 0.742\,\text{m}^3
= 742\,\text{L}.
\]

\subsection{The fill limit and the tank size}

A propane tank is never filled to the brim. The liquid expands as it
warms, and a tank filled liquid-full at a cold temperature would go
hydraulically solid on a hot day, after which any further warming is
resisted by the liquid's bulk modulus rather than by a compressible
vapour space, and the pressure climbs along a slope steeper than the
saturation curve. The standard rule fills to no more than $80\,\%$ of the
tank volume with liquid at the fill temperature, reserving $20\,\%$ as
vapour ullage. The tank volume must therefore be
\[
V_{\text{tank}} = \frac{V_{\text{liq}}}{0.80} = \frac{742}{0.80} = 928\,
\text{L}.
\]
The next standard residential size above this is the nominal
$1000\,\text{L}$ ($\approx 250\,\text{US gallon}$) tank, which we
specify. Figure~\ref{fig:vol04ch02:lpg-tank-fill} carries the fill
fraction of this $80\,\%$ charge across the weather band: the liquid
volume fraction rises from $80\,\%$ at the $15\,\degC$ fill point to
about $89\,\%$ at $50\,\degC$, staying clear of the liquid-full line.
The same figure shows the hazard of overfilling: a charge filled to
$95\,\%$ at $15\,\degC$ reaches liquid-full near $33\,\degC$, an ordinary
summer afternoon, after which the relief valve is the only thing between
the tank and a hydraulic rupture. The $80\,\%$ rule is the engineering
margin that keeps the tank pressure on the saturation curve rather than
on the liquid bulk modulus.

\subsection{The pressure the tank holds across the weather band}

The pressure in the tank is the saturation pressure of propane at the
liquid temperature, read off the saturation curve of
figure~\ref{fig:vol04ch02:propane-vapour-pressure}. At the cold design
point of $-10\,\degC$ the saturation pressure is $345\,\text{kPa}$
absolute; at the mild $25\,\degC$ it is $952\,\text{kPa}$; on a
$50\,\degC$ hot day with the tank in direct sun it is $1713\,\text{kPa}$.
The pressure is set by temperature alone and does not depend on how full
the tank is, because as long as both phases are present the vapour sits at
its saturation pressure. The tank's working-pressure rating must cover
the highest foreseeable storage temperature with margin. Taking the
hot-day liquid temperature at $50\,\degC$ and adding margin for solar
loading on the vapour-space wall, the rating is conventionally set near
$2.5\,\text{MPa}$, with a pressure-relief valve set near $2.07\,\text{MPa}$,
the dashed line in figure~\ref{fig:vol04ch02:propane-vapour-pressure}.
The relief set point is reached not by weather but by a fire-driven wall
temperature, which is the case we size next.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-propane-vapour-pressure.tex}

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-lpg-tank-fill.tex}

\subsection{The vapour off-take rate on the coldest day}

A propane tank delivers gas by boiling liquid, and the boiling draws its
latent heat from the liquid and ultimately from the surroundings through
the tank wall. The coldest-day load of $12\,\text{kW}$ at the furnace,
divided by the same furnace efficiency, demands a fuel power of
$12/0.90 = 13.3\,\text{kW}$, hence a vapourisation rate of
\[
\dot{m} = \frac{P_{\text{fuel}}}{H} = \frac{13.3\times 10^3}{46\times
10^6} = 2.9\times 10^{-4}\,\text{kg}/\text{s} = 1.04\,\text{kg}/\text{h}.
\]
The latent heat this draws is $\dot{m}\, h_{fg}$ with $h_{fg} \approx
358\,\text{kJ}/\text{kg}$ for propane at this temperature, that is
$(2.9\times 10^{-4})(358\times 10^3) = 104\,\text{W}$ pulled out of the
liquid. If the tank wall cannot conduct this in from the cold outside
air fast enough, the liquid cools below ambient, its saturation pressure
falls, and the regulator can starve, the failure mode of an undersized
tank in deep cold. The off-take capacity scales with the wetted wall
area, which is why the cold-climate sizing rule sets a minimum tank
surface even when the monthly volume would allow a smaller vessel; our
$1000\,\text{L}$ tank has ample wetted area for a $1\,\text{kg}/\text{h}$
draw, so the volume sizing governs and the off-take check passes.

\subsection{Relief sizing for the fire case}

The governing load on the relief valve is not weather but external fire.
A tank engulfed in fire receives a heat flux on the order of
$q'' \approx 40\,\text{kW}/\text{m}^2$ over its wetted surface. For the
$1000\,\text{L}$ tank, a horizontal cylinder of roughly $A_{\text{wet}}
\approx 4\,\text{m}^2$ wetted area, the fire input is
\[
\dot{Q}_{\text{fire}} = q''\, A_{\text{wet}} = (40\times 10^3)(4) =
1.6\times 10^5\,\text{W} = 160\,\text{kW}.
\]
All of this heat boils liquid, and the relief valve must pass the vapour
generated, else the pressure runs past the rating. The required vapour
mass rate is
\[
\dot{m}_{\text{relief}} = \frac{\dot{Q}_{\text{fire}}}{h_{fg}} =
\frac{1.6\times 10^5}{358\times 10^3} = 0.45\,\text{kg}/\text{s},
\]
some $1600\,\text{kg}/\text{h}$, more than a thousand times the
coldest-day off-take. The relief valve is sized for this fire-driven
flow, not for the normal draw, which is why relief orifices on
liquefied-gas tanks are large out of all proportion to the service
flow. The vapour leaves at the relief set pressure of $2.07\,\text{MPa}$
and a corresponding saturation temperature near $58\,\degC$; the standard
relief-area formula then sets the orifice from this mass rate, the set
pressure, and the propane vapour properties.

\subsection{The BLEVE margin}

The relief valve protects against overpressure, but it does not protect
the metal. In a sustained fire the vapour-space wall, which the boiling
liquid does not cool, climbs toward the fire temperature and loses
strength, while the relief valve is still passing vapour and holding the
pressure at the set point. The combination, a wall weakening under a
pressure the relief valve is by design keeping high, is the BLEVE
precursor. We size the consequence to fix the margin. The tank holds
$742\,\text{L}$ of liquid, $376\,\text{kg}$, at the moment of a full
fill. If the wall tears and the liquid flashes, the vapour volume at
ambient is
\[
V_{\text{vapour}} = \frac{m\, R_s T}{p_{\text{atm}}} = \frac{(376)(188.5)
(298)}{101{,}325} = 208\,\text{m}^3,
\]
a $280$-fold expansion of the $0.74\,\text{m}^3$ liquid, and if it
ignites, the heat of combustion of the inventory, $m H = (376)(46\times
10^6) = 1.7\times 10^{10}\,\text{J}$, releases in a fireball, equivalent
in energy to several tonnes of TNT. The engineering margin against this
is not a single number but a layered defence: the $80\,\%$ fill limit
keeps a vapour space so the relief valve sees vapour rather than liquid;
the relief valve passes the fire-driven boil-off; a water-deluge or
fire-rated insulation on the vapour-space wall keeps the metal below its
strength-loss temperature for the duration of a credible fire; spacing
from the dwelling and from ignition sources limits both the fire exposure
and the consequence; and a remote shutoff lets responders isolate the
supply. Each layer addresses a different leg of the same chain that the
failure section names.

\subsection{Verdict}

The specification holds. A nominal $1000\,\text{L}$ tank filled to
$80\,\%$ carries the $376\,\text{kg}$ monthly charge with ullage to
spare, holds a working pressure that the saturation curve caps at
$1.7\,\text{MPa}$ on the hottest credible day against a $2.5\,\text{MPa}$
rating, delivers the $1\,\text{kg}/\text{h}$ coldest-day off-take from
its wetted wall area, and carries a relief valve sized for the
$0.45\,\text{kg}/\text{s}$ fire boil-off rather than the service flow.
The honest engineering reading is that the volume sizing governs in this
climate and the off-take check is comfortable, but a colder climate or a
larger peak load would push the off-take rate against the wetted-area
limit and might force a larger tank than the volume alone requires, or a
two-tank manifold to add wall area. The case study has carried one set of
numbers from a heating duty to a signed tank specification; the
discipline, not the particular numbers, is the transferable result. Every
pressurised-liquid vessel clears service by some version of this same
chain: duty, charge, fill limit, working pressure from the saturation
curve, off-take from the wall area, relief from the fire case, and a
consequence margin against the phase boundary the failure section maps.

\section{A second end-to-end case study: a steam Rankine
power-cycle state-point analysis}\pathtag{standard}

The propane case sized a vessel that stores a phase change at rest. A
steam power plant runs a phase change in a loop, twice per cycle, and
turns the latent heat of water into shaft work. We take a subcritical
steam Rankine cycle of the kind that produces most of the world's
grid electricity and carry it end to end: fix the four state points from
the steam tables, balance the energy across each component, compute the
thermal efficiency, and read the result against the Carnot bound and the
turbine-exit moisture limit. Every tool of the chapter appears once. The
saturation tables fix the condenser and boiler-drum states, the
superheat tables fix the turbine inlet, the quality and the lever rule
fix the wet turbine exit, the incompressibility approximation sizes the
pump work, and the energy balance is the first law of chapter~1 applied
component by component. The cycle is the same loop the $T$-$s$ diagram of
figure~\ref{fig:vol04ch02:ts-dome-cycle} sketched schematically, now
carrying real numbers in figure~\ref{fig:vol04ch02:rankine-ts-cycle}.

\subsection{The cycle and its four states}

The plant runs between a condenser at $10\,\text{kPa}$ and a boiler at
$8\,\text{MPa}$, with the steam superheated to $500\,\degC$ at the
turbine inlet. The four states of the ideal cycle are: state~1, the
saturated liquid leaving the condenser; state~2, the compressed liquid
the feed pump delivers to the boiler; state~3, the superheated steam at
the turbine inlet; state~4, the wet steam leaving the turbine. We take
the turbine and pump as isentropic for the ideal analysis and return to
the real-machine corrections at the end. The mass flow rate is
$\dot{m} = 50\,\text{kg}/\text{s}$, a small utility unit.

\subsection{State 1 and the pump work to state 2}

State~1 is saturated liquid at the condenser pressure of $10\,\text{kPa}$.
From the saturated-water table at $10\,\text{kPa}$ the saturation
temperature is $45.8\,\degC$, and
\[
h_1 = h_f = 191.8\,\text{kJ}/\text{kg}, \quad
v_1 = v_f = 1.010\times 10^{-3}\,\text{m}^3/\text{kg}, \quad
s_1 = s_f = 0.649\,\text{kJ}/(\text{kg}\cdot\si{\kelvin}).
\]
The feed pump raises the pressure from $10\,\text{kPa}$ to $8\,\text{MPa}$.
Liquid water is nearly incompressible, so the reversible pump work per
kilogram is the incompressible-flow result $w_p = v_1(p_2 - p_1)$ from
the relation $dh = v\,dp$ at constant entropy and constant $v$:
\[
w_p = v_1(p_2 - p_1) = (1.010\times 10^{-3})(8000 - 10)\times 10^3 =
8.07\times 10^3\,\text{J}/\text{kg} = 8.1\,\text{kJ}/\text{kg}.
\]
The compressed-liquid enthalpy at state~2 is then
\[
h_2 = h_1 + w_p = 191.8 + 8.1 = 199.9\,\text{kJ}/\text{kg},
\]
and the temperature rise across the pump is tiny, $w_p/c \approx 8.1/4.18
= 1.9\,\si{\kelvin}$, which is why states~1 and~2 sit almost on top of one
another on the left limb of the dome in
figure~\ref{fig:vol04ch02:rankine-ts-cycle}. The incompressibility
approximation earns its place here: computing $h_2$ from the
compressed-liquid tables would return the same value to within the table's
own rounding, and the $v\,\Delta p$ shortcut saves a table look-up at
every flowsheet pass.

\subsection{State 3 at the boiler exit}

State~3 is superheated steam at $8\,\text{MPa}$ and $500\,\degC$. The
superheated-steam table at this pressure and temperature gives
\[
h_3 = 3399\,\text{kJ}/\text{kg}, \quad
s_3 = 6.727\,\text{kJ}/(\text{kg}\cdot\si{\kelvin}).
\]
The boiler carries the working fluid from the compressed liquid at
state~2 all the way to this superheated state, three processes in one
constant-pressure path: it warms the liquid to the $8\,\text{MPa}$
saturation temperature of $295\,\degC$, boils it across the two-phase
region at constant temperature, and superheats the dry vapour to
$500\,\degC$. The heat added per kilogram is the enthalpy rise,
\[
q_{\text{in}} = h_3 - h_2 = 3399 - 199.9 = 3199\,\text{kJ}/\text{kg}.
\]

\subsection{State 4 at the turbine exit, and the moisture check}

The turbine expands the steam isentropically from $8\,\text{MPa}$ to the
condenser pressure of $10\,\text{kPa}$, so the exit entropy equals the
inlet entropy, $s_4 = s_3 = 6.727\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$.
At $10\,\text{kPa}$ the saturated entropies are $s_f = 0.649$ and $s_g =
8.150\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$. Since $s_f < s_4 < s_g$,
the exit is a wet two-phase mixture, and the quality follows from
inverting the entropy mixture rule,
\[
x_4 = \frac{s_4 - s_f}{s_g - s_f} = \frac{6.727 - 0.649}{8.150 - 0.649} =
\frac{6.078}{7.501} = 0.810.
\]
The exit enthalpy is then $h_4 = h_f + x_4 h_{fg}$ with $h_f = 191.8$ and
$h_{fg} = 2392.8\,\text{kJ}/\text{kg}$ at $10\,\text{kPa}$,
\[
h_4 = 191.8 + 0.810\times 2392.8 = 191.8 + 1938 = 2130\,\text{kJ}/\text{kg}.
\]
The exit quality of $0.810$ means $19\,\%$ of the stream is liquid water
by mass at the last turbine stage, above the conventional $0.88$ to
$0.90$ moisture limit that protects the low-pressure blades from droplet
erosion. The cycle as drawn would erode its turbine, and the standard
remedies are the ones the worked turbine example named: raise the
superheat, or split the expansion and reheat between stages. Reheat to
$500\,\degC$ at an intermediate pressure of around $1\,\text{MPa}$ would
lift the final quality past $0.92$ and is the reason every utility
Rankine plant of this pressure runs with reheat; chapter~3 of this
volume develops the reheat geometry. We carry the simple cycle forward to
keep the accounting transparent and flag the moisture as the design
defect it is.

\subsection{The energy balance and the efficiency}

The turbine work per kilogram is the enthalpy drop across it,
\[
w_t = h_3 - h_4 = 3399 - 2130 = 1269\,\text{kJ}/\text{kg},
\]
against which the pump consumed $w_p = 8.1\,\text{kJ}/\text{kg}$, so the
net work is
\[
w_{\text{net}} = w_t - w_p = 1269 - 8.1 = 1261\,\text{kJ}/\text{kg}.
\]
The heat rejected at the condenser closes the balance,
\[
q_{\text{out}} = h_4 - h_1 = 2130 - 191.8 = 1938\,\text{kJ}/\text{kg},
\]
and the first-law check holds to table rounding, $q_{\text{in}} -
q_{\text{out}} = 3199 - 1938 = 1261\,\text{kJ}/\text{kg} =
w_{\text{net}}$. The thermal efficiency is the net work divided by the
heat in,
\[
\eta_{\text{th}} = \frac{w_{\text{net}}}{q_{\text{in}}} = \frac{1261}{3199}
= 0.394,
\]
just under $40\,\%$. At the $50\,\text{kg}/\text{s}$ mass flow the gross
power is $\dot{m}\,w_{\text{net}} = (50)(1261) = 6.3\times 10^4\,\text{kW}
= 63\,\text{MW}$, the boiler duty is $\dot{m}\,q_{\text{in}} = 160\,
\text{MW}$, and the condenser must reject $\dot{m}\,q_{\text{out}} = 97\,
\text{MW}$ to the cooling water, the dominant heat flow that sets the
cooling-tower or river-intake size. Figure~\ref{fig:vol04ch02:rankine-energy-balance}
draws the per-kilogram flows as a bar chart and makes the dominant fact
of a steam plant plain: most of the heat added at the boiler leaves at
the condenser, and the pump work is a rounding error against the turbine
work.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-rankine-ts-cycle.tex}

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-rankine-energy-balance.tex}

\subsection{The Carnot bound and the real-machine corrections}

The Carnot efficiency for heat exchange between the boiler-drum
saturation temperature ($295\,\degC = 568\,\si{\kelvin}$) and the
condenser temperature ($45.8\,\degC = 319\,\si{\kelvin}$) is
\[
\eta_{\text{Carnot}} = 1 - \frac{T_{\text{cold}}}{T_{\text{hot}}} = 1 -
\frac{319}{568} = 0.438.
\]
The ideal Rankine cycle reaches $0.394$, that is $90\,\%$ of the Carnot
bound computed at the drum temperature. The gap to Carnot is mostly the
boiler's own irreversibility: heat enters the cycle across a finite
temperature difference while the feedwater warms from $46\,\degC$ to the
$295\,\degC$ saturation temperature, so the average temperature of heat
addition is well below the peak, and the superheat to $500\,\degC$ only
partly recovers it. A real plant falls further below the ideal number. An
isentropic turbine efficiency of $0.85$ would cut the turbine work to
$w_t = 0.85\times 1269 = 1079\,\text{kJ}/\text{kg}$ and lift the exit
quality (the irreversible expansion ends warmer and drier), dropping the
cycle efficiency to about $0.33$; boiler-stack losses, condenser
sub-cooling, and pump inefficiency trim it further. The honest reading is
that a simple subcritical Rankine cycle of this pressure delivers a
real-plant efficiency in the low-to-mid thirties of a percent, that
reheat and regenerative feedwater heating are the moves that recover
several points of it, and that the condenser heat rejection, not the
chimney, carries the bulk of the energy the cycle does not convert.

\subsection{Verdict}

The state-point analysis holds and the cycle's defect is explicit. The
four states close a consistent energy balance, the thermal efficiency of
$0.394$ sits at $90\,\%$ of the Carnot bound taken at the drum
temperature, and the $63\,\text{MW}$ net output follows from the mass
flow. The signing engineer would not release this cycle as drawn: the
turbine-exit quality of $0.810$ is below the blade-erosion limit, and the
fix is reheat or higher superheat, not a tweak to the operating
pressures. The transferable result is the procedure, the same loop the
refrigeration example traced in the opposite direction: fix the saturated
end states from the pressure tables, fix the superheated inlet from its
table, carry entropy through the isentropic legs to find the wet exit
quality, size the pump work from the incompressible $v\,\Delta p$, and
let the first law close component by component. Every steam plant, every
organic-Rankine waste-heat unit, and every refrigeration loop clears its
first design pass by exactly this chain of table look-ups and energy
balances.

\section{A third end-to-end case study: a carbon-dioxide transcritical
refrigeration cycle}\pathtag{standard}

The R-134a worked cycle and the Rankine case both ran their high-side heat
exchange inside the two-phase dome, condensing or boiling at constant
temperature along a saturation isobar. A carbon-dioxide refrigeration
cycle cannot, and the reason is a single number on the phase diagram: the
critical temperature of carbon dioxide is $31\,\degC$, barely above a warm
room and well below the temperature a gas cooler can reject heat to on a
summer day. Above that temperature carbon dioxide has no saturation curve
to condense along, so a CO\textsubscript{2} unit rejecting heat to
$35\,\degC$ ambient runs its high side as a supercritical fluid. We carry
such a cycle end to end: place the four states, see why the high-side
pressure becomes a free design variable, find the pressure that maximises
the coefficient of performance, and read the discharge-pressure trade that
distinguishes a transcritical machine from every subcritical one. The
substance is the case. Carbon dioxide returned to refrigeration as a
natural, non-flammable, low-global-warming working fluid after the
synthetic refrigerants fell out of favour, and the transcritical cycle is
the price its low critical point exacts.

\subsection{Why the cycle goes transcritical}

The critical point of carbon dioxide sits at $T_c = 31.0\,\degC$ and $p_c =
7.38\,\text{MPa}$ (\texttt{data/critical\_constants.csv}). A heat-rejection
side that must dump heat to ambient air at $35\,\degC$ needs a working-fluid
temperature above $35\,\degC$ to drive the heat out, and there is no
two-phase state of carbon dioxide at that temperature: the saturation curve
ends at $31\,\degC$. The high side therefore operates above the critical
pressure, where carbon dioxide is a single supercritical phase that cools
without condensing. The heat exchanger on the high side is called a
\engterm{gas cooler} rather than a condenser for exactly this reason, the
fluid never crosses the dome. The low side is conventional: the evaporator
runs at a subcritical pressure inside the dome, boiling liquid
carbon dioxide at, say, $5\,\degC$ to absorb the refrigeration load.
Figure~\ref{fig:vol04ch02:co2-transcritical-cycle} draws the loop on the
$p$-$h$ plane, with the high-side isobar lying entirely above the critical
pressure and the cold side inside the dome.

\subsection{The four states}

We take an evaporator at $5\,\degC$, where saturated carbon dioxide sits at
a pressure of about $4.0\,\text{MPa}$, and a gas cooler that cools the
high-side fluid to $35\,\degC$ at a high-side pressure we leave as a
variable $p_{\text{gc}}$ for now. The state enthalpies below are read from a
carbon-dioxide property table or from a reference equation of state
(CoolProp, REFPROP), the kind of look-up the chapter's properties section
described; we quote rounded values to carry the accounting.

State~1 is saturated vapour leaving the evaporator at $4.0\,\text{MPa}$
($5\,\degC$): $h_1 = 432\,\text{kJ}/\text{kg}$, $s_1 = 1.74\,\text{kJ}/
(\text{kg}\cdot\si{\kelvin})$. State~2 is the compressor discharge after an
isentropic lift to the gas-cooler pressure; for $p_{\text{gc}} =
9.5\,\text{MPa}$ the isentropic discharge is $h_2 = 478\,\text{kJ}/\text{kg}$
at about $78\,\degC$, deep in the supercritical region. State~3 is the
gas-cooler exit at $9.5\,\text{MPa}$ and $35\,\degC$; the supercritical
fluid there is dense and cold, $h_3 = 253\,\text{kJ}/\text{kg}$. State~4 is
the throttle exit back at the evaporator pressure of $4.0\,\text{MPa}$,
isenthalpic so $h_4 = h_3 = 253\,\text{kJ}/\text{kg}$, a two-phase mixture
of quality
\[
x_4 = \frac{h_4 - h_f}{h_g - h_f} = \frac{253 - 192}{432 - 192} =
\frac{61}{240} = 0.25,
\]
using the saturated-liquid and saturated-vapour enthalpies $h_f =
192\,\text{kJ}/\text{kg}$ and $h_g = 432\,\text{kJ}/\text{kg}$ at the
$4.0\,\text{MPa}$ evaporator pressure. A quarter of the stream flashes
across the throttle, the same flash that drove the R-134a valve, and the
remaining three-quarters of the latent heat is the evaporator's to absorb.

\subsection{Cooling, work, and the COP at one pressure}

The refrigerating effect per kilogram is the evaporator enthalpy rise,
\[
q_L = h_1 - h_4 = 432 - 253 = 179\,\text{kJ}/\text{kg},
\]
and the compressor work per kilogram is the discharge-minus-suction
enthalpy,
\[
w_c = h_2 - h_1 = 478 - 432 = 46\,\text{kJ}/\text{kg}.
\]
The coefficient of performance at this gas-cooler pressure is
\[
\text{COP} = \frac{q_L}{w_c} = \frac{179}{46} = 3.9.
\]
The heat rejected at the gas cooler closes the balance, $q_H = h_2 - h_3 =
478 - 253 = 225\,\text{kJ}/\text{kg}$, and the first law checks: $q_H - q_L
= 225 - 179 = 46\,\text{kJ}/\text{kg} = w_c$. At a refrigeration duty of
$10\,\text{kW}$ the mass flow is $\dot{m} = q_L^{-1}\times 10\,\text{kW} =
10/179 = 0.056\,\text{kg}/\text{s}$ and the compressor draws $\dot{m}\,w_c =
(0.056)(46) = 2.6\,\text{kW}$.

\subsection{The discharge-pressure optimum}

The number that makes a transcritical cycle different is the freedom in the
gas-cooler pressure. In a subcritical machine the condenser pressure is the
saturation pressure at the condensing temperature, fixed by the ambient the
condenser rejects to; the designer has no separate pressure knob. Above the
critical point there is no saturation curve, so the high-side pressure and
the gas-cooler outlet temperature are independent, and the gas-cooler exit
enthalpy $h_3$ depends on both. Raising $p_{\text{gc}}$ at a fixed outlet
temperature pushes the supercritical state into a denser region with a lower
enthalpy, which increases the refrigerating effect $q_L = h_1 - h_3$; it
also raises the compressor discharge enthalpy $h_2$, which increases the
work $w_c$. The two effects pull the COP in opposite directions, and a
maximum sits between them.

Figure~\ref{fig:vol04ch02:co2-gascooler-optimum} traces the COP against
$p_{\text{gc}}$ for the $35\,\degC$ outlet. At a low high-side pressure of
$8.0\,\text{MPa}$, just above critical, the gas-cooler exit is warm and
enthalpy-rich, the refrigerating effect is small, and the COP is only about
$2.0$. Raising the pressure to $9.5\,\text{MPa}$ drops $h_3$ steeply,
$q_L$ rises faster than $w_c$, and the COP climbs to its maximum near $2.9$
to $3.9$ depending on the evaporator temperature and compressor efficiency.
Past the optimum the exit enthalpy keeps falling, but only slowly, while the
compressor work keeps climbing, so the COP declines: at $13\,\text{MPa}$ it
has fallen back to about $2.1$. The optimal gas-cooler pressure follows a
near-linear rule of thumb with the gas-cooler outlet temperature,
$p_{\text{gc,opt}}\,[\text{bar}] \approx 2.7\,T_{\text{gc,out}}\,[\degC] -
9.6$, which for a $35\,\degC$ outlet gives about $85\,\text{bar}$
($8.5\,\text{MPa}$) to $95\,\text{bar}$, the neighbourhood the figure marks.
A real CO\textsubscript{2} unit carries a high-side pressure controller that
holds the discharge at this optimum as the ambient and the load move,
because, unlike a condenser pressure that settles itself, the transcritical
high-side pressure must be actively driven to where the COP peaks.

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-co2-transcritical-cycle.tex}

\input{volumes/04-energy/ch02-equations-of-state-phase-changes/figures/fig-co2-gascooler-optimum.tex}

\subsection{The gas-cooler outlet temperature matters more than anything}

The outlet temperature of the gas cooler sets the floor on the cycle. A
gas cooler that can reach $35\,\degC$ delivers $h_3 = 253\,\text{kJ}/
\text{kg}$ and a peak COP near $3$; a gas cooler limited to $40\,\degC$ on a
hotter day leaves $h_3$ higher, around $280\,\text{kJ}/\text{kg}$, cutting
the refrigerating effect to $q_L = 432 - 280 = 152\,\text{kJ}/\text{kg}$ and
dropping the COP by a fifth even at the optimised pressure. The strong
sensitivity of $h_3$ to the gas-cooler approach temperature is the reason
transcritical CO\textsubscript{2} systems perform best in cool climates and
why supermarket installations in northern Europe favour them while warm
climates need an added internal heat exchanger, parallel compression, or an
ejector to recover the throttling loss. The cycle's weakness is the same
flash the R-134a throttle showed, made worse: the supercritical-to-two-phase
expansion across a simple valve discards a larger fraction of the available
work than a subcritical throttle, which is why CO\textsubscript{2} systems
are the ones that most often replace the valve with an expander or an
ejector.

\subsection{Verdict}

The transcritical cycle holds and its defining trade is explicit. The four
states close a consistent energy balance, the COP of $3.9$ at the
$9.5\,\text{MPa}$ optimum and the $5\,\degC$ evaporator is competitive with
a synthetic-refrigerant cycle in a cool climate, and the $10\,\text{kW}$
duty sizes a $0.056\,\text{kg}/\text{s}$ flow and a $2.6\,\text{kW}$
compressor. The transferable result is the discharge-pressure trade: in a
transcritical cycle the high-side pressure is a control variable with a true
optimum, not a quantity a condensing saturation curve hands the designer for
free. A CO\textsubscript{2} unit that runs its gas cooler at a fixed
pressure leaves performance on the table at every ambient but the one it was
tuned for, and the working discipline is to track the optimal high-side
pressure with the gas-cooler outlet temperature. The procedure, fix the four
states, find $q_L$ and $w_c$ from the enthalpy differences, and sweep the
free high-side pressure for the COP maximum, carries to every transcritical
machine and to any cycle whose high side rejects heat above the working
fluid's critical temperature.

\section{Failure: cavitation, water-hammer, BLEVE}\pathtag{core}

Phase transitions are the source of three working classes of failure
in fluid and thermal systems. The chapter's failure section names the
mechanisms; the named industrial accidents are catalogued in Volume X.

\subsection{Cavitation}

Cavitation is the formation of vapour bubbles in a liquid at locations
where the local pressure drops below the saturation pressure at the
liquid's local temperature. The bubbles form at points of high local
velocity (the suction side of a pump impeller, the low-pressure side
of a propeller blade, the throat of a venturi, the inside of a sharp
elbow) and collapse violently when the liquid returns to higher
pressure. The collapse generates microjets of liquid that strike the
nearby solid surface at velocities approaching $100\,\text{m}/\text{s}$,
producing pitting and progressive material loss.

Cavitation has three engineering consequences. The first is the loss
of pump capacity: a cavitating pump delivers less head than its
non-cavitating performance curve predicts, because the vapour bubbles
in the impeller passage reduce the effective fluid density. The second
is mechanical damage: cavitation pitting on pump impellers, ship
propellers, and turbine blades is the primary erosion mode in
high-speed liquid machinery. The third is noise and vibration: the
collapse of vapour cavities produces broadband acoustic emission that
is the diagnostic signal of incipient cavitation, audible long before
the damage accumulates.

The standard engineering parameter is the \engterm{Net Positive
Suction Head} (NPSH), the margin in pressure head above the saturation
pressure that the pump's suction must maintain to avoid cavitation.
Pump manufacturers specify the required NPSH ($\text{NPSH}_r$) for
each operating point; the installation must provide at least that much
available NPSH ($\text{NPSH}_a$, computed from the suction-side
geometry, the fluid's vapour pressure at the operating temperature,
and any frictional losses upstream). A pump installation with
$\text{NPSH}_a < \text{NPSH}_r$ will cavitate, with diagnostic noise
and progressive damage.

The available NPSH is a head balance at the pump suction, and writing it out
shows which term the operating temperature attacks. Referred to the pump
centreline,
\[
\text{NPSH}_a = \frac{p_{\text{atm}} - p_{\text{sat}}(T)}{\rho g} +
z_{\text{stat}} - h_{\text{friction}},
\]
the sum of the atmospheric-minus-vapour head, the static head of the source
above the pump, and a debit for suction-line friction. The vapour-pressure term
is the one the saturation curve governs, and it is the term that fails a hot
service. For cold water at $20\,\degC$ the saturation pressure is $2.3\,\text{kPa}$,
so the atmospheric-minus-vapour head is $(101.3 - 2.3)\times 10^3/(1000\times
9.81) = 10.1\,\text{m}$, nearly the full atmospheric head; for water at
$90\,\degC$ the saturation pressure has climbed to $70\,\text{kPa}$ and the same
head has collapsed to $(101.3 - 70)\times 10^3/(1000\times 9.81) = 3.2\,\text{m}$,
a two-thirds loss of margin bought by nothing but the temperature. The steep
rise of $p_{\text{sat}}(T)$ along the saturation curve is the whole story of why
boiler-feed and hot-condensate pumps are the ones that cavitate, and why such
pumps are mounted low, fed from an elevated deaerator, or run with a booster
that adds static head. The margin can be nondimensionalised as the
\engterm{Thoma cavitation number} $\sigma = \text{NPSH}_a/H$, the ratio of the
suction margin to the pump's developed head, and each pump has a critical
$\sigma$ below which its head begins to break down; matching the installation's
$\sigma$ to the manufacturer's critical value is the design check that sits
behind the raw $\text{NPSH}_a \ge \text{NPSH}_r$ inequality. The temperature
sensitivity also explains a field diagnostic: a pump that runs quietly on a cold
morning and rattles by afternoon is not failing mechanically, it is watching its
vapour-pressure head shrink as the fluid warms, and the cure is to restore
suction margin rather than to overhaul the machine.

\subsection{Water-hammer}

Water-hammer is the pressure transient that results when a flowing
liquid is rapidly decelerated, typically by a valve closure. The
deceleration converts the liquid's kinetic energy into elastic strain
energy in the liquid and the pipe wall, raising the pressure at the
deceleration point. The pressure pulse propagates back upstream at the
speed of sound in the liquid-filled pipe ($\sim 1000$ to $1400\,
\text{m}/\text{s}$ in steel pipes filled with water), reflects from
the upstream boundary, and continues to oscillate until damped.

The peak pressure rise from a sudden valve closure, in the linear
acoustic limit, is given by the \engterm{Joukowsky equation}
\[
\Delta p = \rho c \Delta v,
\]
with $\rho$ the liquid density, $c$ the wave speed in the pipe-fluid
system, and $\Delta v$ the velocity change. For water flowing at
$2\,\text{m}/\text{s}$ in a steel pipe with $c = 1200\,\text{m}/
\text{s}$, the peak pressure rise is $\rho c \Delta v = 1000 \times
1200 \times 2 = 2.4\,\text{MPa}$, comparable to the working pressure
of most water-supply piping. A pipe rated for $1\,\text{MPa}$ steady-
state working pressure can rupture when subjected to a
water-hammer pulse of this magnitude.

The engineering control of water-hammer involves slow valve closure
(closure time longer than the round-trip wave-travel time $2L/c$),
air chambers (volumes of compressible gas in the system that absorb
the pressure pulse), surge tanks (open columns of water that absorb
the pressure pulse by raising and lowering their water level), and
pressure-relief valves set to open at the predicted maximum surge
pressure. Water-hammer accidents in industrial steam systems,
hydropower penstocks, and municipal water mains are catalogued in
Volume X.

\subsection{BLEVE}

A BLEVE (Boiling Liquid Expanding Vapour Explosion) is the catastrophic
failure of a pressurised vessel containing a liquid at a temperature
above its atmospheric-pressure boiling point. The vessel may be heated
externally (a propane tank engulfed in fire) or the failure may begin
mechanically (a tank rupture due to corrosion, impact, or
manufacturing defect). When the vessel ruptures, the pressure on the
liquid drops to atmospheric, the liquid flashes to vapour explosively
(typical specific-volume change of $200\times$ to $300\times$), and
the resulting vapour cloud may ignite if the contents are flammable.

The energy released in a BLEVE comes from three sources: the
thermal energy stored in the superheated liquid (the liquid's
temperature, minus its atmospheric boiling point, times its mass times
the specific heat); the volume work of the flashing vapour against
atmospheric pressure; and, if the contents are flammable and ignite,
the heat of combustion of the entire flammable inventory in a fireball.

Industrial BLEVE accidents include the Feyzin refinery explosion in
1966, the Mexico City PEMEX LPG terminal disaster in 1984 (San
Juanico), and the Crescent City Illinois propane railway tank-car
explosions in 1970. Each followed the same sequence: external fire on
a pressurised liquefied-gas tank, weakening of the metal at the
vapour-space wall (which receives less cooling than the liquid-wetted
wall), tank rupture, flashing of the liquid, and ignition of the
expanding vapour cloud.

The engineering control of BLEVE involves: thermal protection of the
tank wall (insulation, sprinkler systems that water-cool the tank
under fire conditions); secondary containment (berms, drainage to
remote impoundment); spacing from other equipment (so a BLEVE in one
tank does not engulf a second); and operational discipline (no fire-
prone activities near pressurised liquefied-gas storage). The chemical-
process-safety standards (OSHA PSM in the United States, Seveso
Directive in the European Union) embed these controls in regulatory
form.

\begin{principle}[Phase transitions are the failure surface]
The two-phase boundary on a substance's $p$-$T$ diagram is the working
failure surface of fluid and thermal systems. Cavitation appears
where flow drives liquid pressure below saturation at the operating
temperature; water-hammer appears where rapid deceleration of a
flowing liquid generates pressure pulses that exceed pipe ratings;
BLEVE appears where a pressurised liquid is exposed to a heat source
that drives it above its atmospheric saturation. The engineering
disciplines for each (NPSH margin, valve-closure timing and surge
protection, BLEVE-resistant tank design) are different in detail and
identical in spirit: stay clear of the phase boundary, or provide for
the consequences of crossing it.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Trace a real refrigerant cycle on a $p$-$h$ diagram]
The reader instruments a working vapour-compression cycle (the
household refrigerator from chapter~1, or a window-unit air
conditioner, or a heat pump) over a quasi-steady operating period,
measures the four cycle-state points, and plots the cycle on the
refrigerant's $p$-$h$ diagram.

\textbf{Track}. Instrumentation plus property-table analysis.
\textbf{Hazard class}: low.

\textbf{Measurements}.

\begin{itemize}[leftmargin=*]
  \item Refrigerant identification: from the appliance nameplate
    (R-134a, R-410A, R-1234yf, or other).
  \item Evaporator pressure: from the low-side service port (if
    accessible) or inferred from the evaporator surface temperature
    via the saturation curve. A non-contact surface thermometer reads
    the cold-side coil at one or more locations.
  \item Condenser pressure: from the high-side service port or
    inferred from the condenser surface temperature similarly.
  \item Compressor inlet temperature: surface thermometer on the
    suction line near the compressor.
  \item Compressor discharge temperature: surface thermometer on the
    discharge line near the compressor.
  \item Throttle inlet temperature: surface thermometer on the
    liquid line near the throttle.
\end{itemize}

\textbf{Analysis}.

\begin{itemize}[leftmargin=*]
  \item Look up the refrigerant's properties at each of the four
    state points using a table or software.
  \item Plot the four state points on a $p$-$h$ diagram for the
    refrigerant. Compare the plotted cycle to the textbook ideal
    cycle (compression isentropic, condensation at constant pressure
    in the two-phase region, throttling isenthalpic, evaporation at
    constant pressure in the two-phase region).
  \item Identify the irreversibilities visible in the plotted cycle:
    sub-cooling at the condenser exit, superheat at the evaporator
    exit, non-isentropic compression, pressure drop across the heat
    exchangers.
  \item Compute the cycle's specific work, specific cooling capacity,
    and COP from the measured state points; compare to the Carnot
    bound for the measured evaporator and condenser temperatures.
\end{itemize}

\textbf{Reflection ($1{,}500$ to $2{,}500$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
  \item The discrepancy between the measured cycle and the textbook
    ideal cycle; the largest source of irreversibility identified.
  \item The effect of the sub-cooling and superheat on the cycle's
    COP.
  \item One design change that would close the gap to the Carnot
    bound, and the cost of that change.
\end{itemize}

\textbf{Deliverable}. The measured-state-point table, the $p$-$h$
diagram with the plotted cycle, the COP calculation, and the
reflection.

\textbf{Time}. Two to four hours of measurement, two to four hours of
chart work, four to six hours of write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A rigid tank of volume $0.05\,\text{m}^3$ contains nitrogen at $300\,
\si{\kelvin}$ and $5\,\text{MPa}$. Treating nitrogen as an ideal gas
with $R_s = 297\,\text{J}/(\text{kg}\cdot\si{\kelvin})$, find the
mass of nitrogen in the tank.

\paragraph{Solution sketch.} Solve $pV = m R_s T$ for the mass:
$m = pV/(R_s T) = (5\times 10^6)(0.05)/(297\times 300) = 250{,}000/89{,}100
= 2.81\,\text{kg}$. Carry the units: $\text{Pa}\cdot\text{m}^3 = \text{J}$
in the numerator, $\text{J}/\text{kg}$ in the denominator, leaving
kilograms.
\end{exercise}

\begin{exercise}
Repeat the previous exercise using the compressibility factor $Z$
from the generalised chart. Nitrogen has $T_c = 126.2\,\si{\kelvin}$
and $p_c = 3.39\,\text{MPa}$. Compute the percent error introduced by
the ideal-gas assumption.

\paragraph{Solution sketch.} The reduced coordinates are $T_r =
300/126.2 = 2.38$ and $p_r = 5/3.39 = 1.47$. At $T_r > 2$ the chart sits
close to $Z \approx 1$, dipping only slightly at this $p_r$; read
$Z \approx 0.99$ to $1.00$. The real mass is $m = pV/(Z R_s T) =
2.81/Z \approx 2.81$ to $2.84\,\text{kg}$, so the ideal-gas estimate is
high by at most about $1\,\%$. The check confirms the qualitative rule:
well above the critical temperature the ideal-gas law is safe even at
$p_r$ of order one. The van der Waals proxy in
\texttt{code/ideal\_vs\_real\_gas.py} returns $Z \approx 0.99$ at these
conditions.
\end{exercise}

\begin{exercise}
Steam at $5\,\text{MPa}$ and $300\,\degC$ enters a turbine and exits
at $0.1\,\text{MPa}$. Using the steam tables, find the inlet and
outlet enthalpies and entropies, and compute the maximum (isentropic)
work output per kilogram.

\paragraph{Solution sketch.} Inlet (superheated, $5\,\text{MPa}$,
$300\,\degC$): $h_1 \approx 2925\,\text{kJ}/\text{kg}$, $s_1 \approx 6.21
\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$. For isentropic expansion to
$0.1\,\text{MPa}$, set $s_2 = s_1 = 6.21$. At $0.1\,\text{MPa}$,
$s_f = 1.30$ and $s_g = 7.36\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$, so
the exit is two-phase with quality $x_2 = (6.21 - 1.30)/(7.36 - 1.30) =
0.81$. Then $h_2 = h_f + x_2 h_{fg} = 417.5 + 0.81\times 2257.5 \approx
2247\,\text{kJ}/\text{kg}$, and $w = h_1 - h_2 \approx 678\,\text{kJ}/
\text{kg}$. The exit quality of $0.81$ is below the $0.90$ blade-erosion
limit, the same warning the worked turbine example raised.
\end{exercise}

\begin{exercise}
Two-phase water at $200\,\degC$ has a quality of $0.4$. Find the
specific volume, the specific enthalpy, and the specific entropy.

\paragraph{Solution sketch.} From the saturated-water table at
$200\,\degC$ (\texttt{data/water\_saturation.csv}): $v_f = 0.001157$,
$v_g = 0.1272\,\text{m}^3/\text{kg}$; $h_f = 852.3$, $h_g = 2792.0\,
\text{kJ}/\text{kg}$; $s_f = 2.331$, $s_g = 6.430\,\text{kJ}/(\text{kg}
\cdot\si{\kelvin})$. Apply $y = (1-x)y_f + x y_g$ with $x = 0.4$:
$v = 0.6(0.001157) + 0.4(0.1272) = 0.0516\,\text{m}^3/\text{kg}$;
$h = 0.6(852.3) + 0.4(2792.0) = 1628\,\text{kJ}/\text{kg}$;
$s = 0.6(2.331) + 0.4(6.430) = 3.97\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$.
\end{exercise}

\begin{exercise}
R-134a at $40\,\degC$ has a saturation pressure of $1017\,\text{kPa}$.
A saturated-liquid sample is throttled from $1017\,\text{kPa}$ to
$200\,\text{kPa}$ (saturation temperature $-10\,\degC$). Find the
quality at the exit of the throttle.

\paragraph{Solution sketch.} Throttling is isenthalpic, so
$h_{\text{out}} = h_f(40\,\degC)$. From the R-134a table
(\texttt{data/r134a\_saturation.csv}), $h_f(40\,\degC) = 107.32\,
\text{kJ}/\text{kg}$. At the exit ($-10\,\degC$), $h_f = 38.55$ and
$h_g = 244.51\,\text{kJ}/\text{kg}$. The quality is $x = (h - h_f)/(h_g -
h_f) = (107.32 - 38.55)/(244.51 - 38.55) = 68.77/205.96 = 0.334$. About a
third of the liquid flashes to vapour across the throttle, the flash that
provides the evaporator's cooling.
\end{exercise}

\begin{exercise}
A piston-cylinder contains $1\,\text{kg}$ of water at $100\,\degC$
with quality $0.6$. Heat is added at constant pressure until the
water becomes saturated vapour. Find the heat required and the work
done.

\paragraph{Solution sketch.} At $100\,\degC$ the pressure is
$101.3\,\text{kPa}$ and stays constant. The process runs from $x_1 = 0.6$
to $x_2 = 1$. For a constant-pressure process the heat equals the
enthalpy change: $q = h_2 - h_1 = (h_g) - (h_f + 0.6 h_{fg}) = (1 -
0.6)h_{fg} = 0.4\times 2256.5 = 902.6\,\text{kJ}/\text{kg}$, so for
$1\,\text{kg}$, $Q = 903\,\text{kJ}$. The work is $W = p(V_2 - V_1) =
p\,m(v_g - v_1)$ with $v_1 = v_f + 0.6(v_g - v_f) = 0.001043 +
0.6(1.6720 - 0.001043) = 1.0036\,\text{m}^3/\text{kg}$ and $v_g = 1.6720$:
$W = 101{,}300\times 1\times(1.6720 - 1.0036) = 67.7\,\text{kJ}$.
\end{exercise}

\begin{exercise}
A propane tank ($T_c = 369.8\,\si{\kelvin}$, $p_c = 4.25\,\text{MPa}$)
holds liquid propane at $25\,\degC$. Find the saturation pressure
and the specific volumes of saturated liquid and saturated vapour
using the generalised chart or the Antoine equation.

\paragraph{Solution sketch.} The reduced temperature is $T_r =
298.15/369.8 = 0.806$. The corresponding-states saturation correlation
(or an Antoine fit) gives $p_r^{\text{sat}} \approx 0.225$, hence
$p_{\text{sat}} \approx 0.225\times 4.25 = 0.96\,\text{MPa}$ (handbook
value $\approx 0.95\,\text{MPa}$ gauge plus atmospheric). The
saturated-liquid specific volume is set by the liquid density, about
$494\,\text{kg}/\text{m}^3$ at $25\,\degC$, so $v_f \approx 2.0\times
10^{-3}\,\text{m}^3/\text{kg}$. Treat the vapour as ideal: $v_g \approx
R_s T/p = (188.5)(298.15)/(0.96\times 10^6) = 0.0586\,\text{m}^3/
\text{kg}$ with $R_s = R/M = 8.314/0.0441$. The large ratio $v_g/v_f
\approx 30$ justifies the $v_g \gg v_f$ approximation used in
Clausius-Clapeyron.
\end{exercise}

\begin{exercise}
Compute the Clausius-Clapeyron prediction for the saturation pressure
of water at $150\,\degC$, given that at $100\,\degC$ the saturation
pressure is $101.325\,\text{kPa}$ and the latent heat is $2257\,
\text{kJ}/\text{kg}$. Compare to the tabulated value.

\paragraph{Solution sketch.} Use the integrated form $\ln(p_2/p_1) =
-(h_{fg}/R_s)(1/T_2 - 1/T_1)$ with $R_s = 461.5\,\text{J}/(\text{kg}
\cdot\si{\kelvin})$, $T_1 = 373.15$, $T_2 = 423.15\,\si{\kelvin}$. The
exponent is $-(2257\times 10^3/461.5)(1/423.15 - 1/373.15) =
-(4890)(2.3633\times 10^{-3} - 2.6799\times 10^{-3}) = -(4890)(-3.166
\times 10^{-4}) = 1.548$. Then $p_2 = 101.325\,e^{1.548} = 101.325\times
4.70 = 476\,\text{kPa}$. The steam-table value is $475.8\,\text{kPa}$,
agreement to better than $1\,\%$ because the latent heat is nearly
constant over this $50\,\si{\kelvin}$ span. The routine
\texttt{code/clausius\_clapeyron.py} reproduces the number.
\end{exercise}

\begin{exercise}
A steel pipe carrying water at $20\,\degC$ has a flow velocity of
$3\,\text{m}/\text{s}$. The downstream valve closes in $0.1\,\text{s}$.
The pipe is $200\,\text{m}$ long. Compute the Joukowsky pressure rise
and determine whether the closure is "rapid" (closure time less than
the wave-round-trip time $2L/c$ with $c \approx 1200\,\text{m}/
\text{s}$).

\paragraph{Solution sketch.} The wave-round-trip time is $2L/c =
2(200)/1200 = 0.333\,\text{s}$. The closure of $0.1\,\text{s} < 0.333\,
\text{s}$, so the closure is rapid and the full Joukowsky rise applies:
$\Delta p = \rho c \Delta v = (1000)(1200)(3) = 3.6\,\text{MPa}$. A
closure slower than $0.333\,\text{s}$ would reduce the surge in roughly
the ratio $2L/(c\,t_{\text{close}})$; to keep the surge to a third of the
Joukowsky value the valve would need about $1\,\text{s}$.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the Clausius-Clapeyron equation $dp/dT = h_{fg}/(T \Delta v)$
from the equality of Gibbs free energies of the two coexisting
phases along the saturation curve.

\paragraph{Solution sketch.} Along the saturation curve the specific
Gibbs free energies of the two phases are equal, $g_f = g_g$, at every
point. Move along the curve by $(dp, dT)$: the equality is preserved, so
$dg_f = dg_g$. With $dg = -s\,dT + v\,dp$ for each phase, $-s_f\,dT +
v_f\,dp = -s_g\,dT + v_g\,dp$. Collect: $(v_g - v_f)\,dp = (s_g -
s_f)\,dT$, hence $dp/dT = (s_g - s_f)/(v_g - v_f)$. Along the isothermal,
isobaric phase change the entropy change is $s_g - s_f = h_{fg}/T$
(reversible heat at constant $T$), giving $dp/dT = h_{fg}/[T(v_g - v_f)] =
h_{fg}/(T\,\Delta v)$.
\end{exercise}

\begin{exercise}
Derive the simplified Clausius-Clapeyron form $\ln p = -h_{fg}/(R_s
T) + C$ from the full equation, stating the two approximations used
(ideal-gas vapour and $v_g \gg v_f$).

\paragraph{Solution sketch.} Start from $dp/dT = h_{fg}/[T(v_g - v_f)]$.
First approximation: $v_g \gg v_f$, so $v_g - v_f \approx v_g$. Second
approximation: the vapour is an ideal gas, $v_g = R_s T/p$. Substituting,
$dp/dT = h_{fg}\,p/(R_s T^2)$, which separates: $dp/p = (h_{fg}/R_s)\,
dT/T^2$. Integrate, treating $h_{fg}$ as constant: $\ln p = -h_{fg}/(R_s
T) + C$. The constant $C$ is fixed by one known $(p, T)$ point on the
saturation curve. The plot of $\ln p$ against $1/T$ is then a straight
line of slope $-h_{fg}/R_s$, as in
figure~\ref{fig:vol04ch02:clausius-clapeyron}.
\end{exercise}

\begin{exercise}
From the van der Waals equation, derive the critical-point conditions
$(\partial p/\partial V)_{T_c} = 0$, $(\partial^2 p/\partial V^2)_{T_c}
= 0$ and show that $a = 27 R^2 T_c^2 / (64 p_c)$ and $b = R T_c /
(8 p_c)$.
\paragraph{Solution sketch.} Write the molar form $p = RT/(v - b) -
a/v^2$. The critical isotherm has a horizontal inflexion, so both the
first and second volume derivatives vanish there:
$(\partial p/\partial v)_T = -RT/(v-b)^2 + 2a/v^3 = 0$ and
$(\partial^2 p/\partial v^2)_T = 2RT/(v-b)^3 - 6a/v^4 = 0$. Divide the two
to eliminate $a$ and $T$: from the ratio one finds $v_c = 3b$. Substitute
back into the first condition to get $T_c = 8a/(27Rb)$, and into the
equation of state to get $p_c = a/(27b^2)$. Inverting the last two for
$a$ and $b$ in terms of $T_c$ and $p_c$ gives $b = RT_c/(8p_c)$ and
$a = 27R^2T_c^2/(64p_c)$. A corollary is $Z_c = p_c v_c/(RT_c) = 3/8 =
0.375$ for every substance, the universal value flagged in the mastery
box.
\end{exercise}

\begin{exercise}
Derive the Joukowsky equation $\Delta p = \rho c \Delta v$ from the
momentum balance across a wave front in a pipe.

\paragraph{Solution sketch.} Take a control volume that moves with the
pressure wave front travelling at speed $c$ into fluid moving at velocity
$v$. In the frame of the front, fluid enters at relative speed $c + v
\approx c$ and the mass flux per unit area is $\rho c$. The front
decelerates the fluid by $\Delta v$. Newton's second law on the control
volume sets the net pressure force per unit area equal to the rate of
momentum change: $\Delta p = (\rho c)\,\Delta v$. Equivalently, the
kinetic energy of the stopped fluid is stored as elastic strain energy in
the fluid and pipe wall, and the wave speed $c$ folds in both the fluid
bulk modulus and the pipe-wall elasticity.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the volume of dry air at sea-level pressure and $20\,\degC$
that contains one mole of nitrogen. State the assumed mole fraction
of nitrogen in air.

\paragraph{Reference estimate.} Take the nitrogen mole fraction of dry
air as $0.78$. One mole of nitrogen therefore comes with $1/0.78 = 1.28$
moles of air total. The molar volume of an ideal gas at
$101.325\,\text{kPa}$ and $293.15\,\si{\kelvin}$ is $v_m = RT/p =
(8.314)(293.15)/(101{,}325) = 0.02406\,\text{m}^3/\text{mol} = 24.1\,
\text{L}/\text{mol}$. The required air volume is $1.28\times 24.1 = 30.8\,
\text{L}$. (At the textbook $0\,\degC$ standard, the molar volume is
$22.4\,\text{L}$ and the answer would be $28.7\,\text{L}$.)
\end{exercise}

\begin{exercise}
Estimate the pressure inside a sealed plastic water bottle when the
ambient temperature rises from $20\,\degC$ to $40\,\degC$. Treat the
air space inside as an ideal gas at constant volume.

\paragraph{Reference estimate.} At constant volume the ideal gas obeys
$p_2/p_1 = T_2/T_1$ in absolute temperature. With $T_1 = 293.15$ and
$T_2 = 313.15\,\si{\kelvin}$, the ratio is $313.15/293.15 = 1.068$.
Starting from atmospheric $p_1 = 101\,\text{kPa}$ absolute, $p_2 = 108\,
\text{kPa}$ absolute, a rise of about $7\,\text{kPa}$ ($0.07\,\text{atm}$,
roughly $1\,\text{psi}$). Modest, which is why a warmed sealed bottle
feels firm but does not burst; the gauge pressure rises only a few
percent because the absolute temperature rises only a few percent. If the
bottle holds liquid as well, evaporation adds to this, the effect the
BLEVE section turns dangerous at higher temperatures.
\end{exercise}

\begin{exercise}
Estimate the volume of saturated vapour produced when $1\,\text{kg}$
of liquid propane at $25\,\degC$ flashes through a regulator to
$1\,\text{atm}$.

\paragraph{Reference estimate.} At $1\,\text{atm}$ propane boils at about
$-42\,\degC$. The flashing is roughly isenthalpic; a fraction of the
liquid evaporates and most ends up as cold vapour after the regulator and
subsequent warming to ambient. Estimate the vapour volume of $1\,\text{kg}$
of propane as ideal gas at ambient $25\,\degC$ and $1\,\text{atm}$:
$n = 1000/44.1 = 22.7\,\text{mol}$, $V = nRT/p = (22.7)(8.314)(298.15)/
(101{,}325) = 0.555\,\text{m}^3$, about $555\,\text{L}$. So a single
kilogram of liquid propane (about $2\,\text{L}$) expands to roughly
$550\,\text{L}$ of gas, a volume ratio near $270\times$, the same order
flagged in the BLEVE discussion.
\end{exercise}

\begin{exercise}
Estimate the cavitation margin (NPSH available) for a household
boiler-feed pump drawing water at $90\,\degC$ from an open tank
located $3\,\text{m}$ above the pump suction. State the assumed
saturation pressure and the suction-line friction loss.

\paragraph{Reference estimate.} Available NPSH in head:
$\text{NPSH}_a = (p_{\text{atm}} - p_{\text{sat}})/(\rho g) + z_{\text{stat}}
- h_{\text{friction}}$. At $90\,\degC$ the saturation pressure of water is
about $70\,\text{kPa}$, so $(101.3 - 70)\times 10^3/(1000\times 9.81) =
3.2\,\text{m}$. The static head is $+3\,\text{m}$ (tank above the pump).
Assume a suction-line friction loss of $1\,\text{m}$. Then
$\text{NPSH}_a \approx 3.2 + 3 - 1 = 5.2\,\text{m}$. The high water
temperature consumes most of the atmospheric head through the large
vapour pressure, which is why hot-water pumps cavitate so readily and
why the tank is placed above the pump to add static head.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a van der Waals equation-of-state solver. Given $T$ and
$v$, return $p$. Plot the isotherms for carbon dioxide ($T_c = 304\,
\si{\kelvin}$, $p_c = 7.4\,\text{MPa}$) for $T = 250, 300, 350\,\si{
\kelvin}$ and identify the two-phase region using the Maxwell
equal-area construction.

\paragraph{Solution sketch.} Compute the carbon-dioxide constants from
the critical data: $a = 27R^2T_c^2/(64p_c) = 0.366\,\text{Pa}\cdot
\text{m}^6/\text{mol}^2$ and $b = RT_c/(8p_c) = 4.28\times 10^{-5}\,
\text{m}^3/\text{mol}$. Evaluate $p(T, v) = RT/(v - b) - a/v^2$ over a
volume grid for each temperature. The $250\,\si{\kelvin}$ isotherm
($T_r = 0.82$) shows the non-monotonic loop; the $300\,\si{\kelvin}$
isotherm ($T_r = 0.99$) is near the critical inflexion; the $350\,
\si{\kelvin}$ isotherm ($T_r = 1.15$) is monotonic. The Maxwell line is
the constant pressure $p_{\text{sat}}$ for which $\int_{v_f}^{v_g}(p -
p_{\text{sat}})\,dv = 0$, found by bisecting on $p_{\text{sat}}$ between
the loop minimum and maximum. The reduced-variable form of the
construction is implemented in \texttt{code/vdw\_isotherms.py}; clamp the
grid to $v > b$ to keep the $1/(v - b)$ singularity off the plot.
\end{exercise}

\begin{exercise}
Simulate a water-hammer transient in a $100\,\text{m}$ horizontal
pipe of internal diameter $50\,\si{\milli\meter}$ when a downstream
valve closes linearly over $0.5\,\text{s}$ in water flowing initially
at $2\,\text{m}/\text{s}$. Use the method of characteristics. Plot
the pressure at the closing valve over the first ten round-trip
times.

\paragraph{Solution sketch.} Discretise the pipe into $N$ reaches and
march the two characteristic equations $C^{+}$ and $C^{-}$ in time, with
the wave speed $c$ setting the Courant condition $\Delta t = \Delta x/c$.
Take $c \approx 1200\,\text{m}/\text{s}$, so the round-trip time is
$2L/c = 0.167\,\text{s}$ and the $0.5\,\text{s}$ closure spans about three
round trips (a slow closure). The valve boundary imposes a closing area
schedule that drives the discharge to zero over $0.5\,\text{s}$. Because
the closure is slower than $2L/c$, the peak pressure stays below the full
Joukowsky value $\rho c \Delta v = (1000)(1200)(2) = 2.4\,\text{MPa}$, by
roughly the factor $2L/(c\,t_{\text{close}}) = 0.33$, giving a peak near
$0.8\,\text{MPa}$. The plotted pressure shows the characteristic stepped
rise during closure followed by decaying oscillation at the period $2L/c$.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify the size and pressure rating of a propane tank for a
household kitchen-stove application requiring $5\,\text{kg}$ of
propane per week, with refill intervals not exceeding one month.
State the saturation pressure assumed at the highest summer
temperature.

\paragraph{Model-answer sketch.} One month at $5\,\text{kg}/\text{week}$
is about $20\,\text{kg}$ of propane. Liquid propane density at storage is
near $0.50\,\text{kg}/\text{L}$, so the liquid volume is $40\,\text{L}$;
with the standard $80\,\%$ fill limit (ullage for thermal expansion of
the liquid) the tank must be at least $50\,\text{L}$. A standard $45$ to
$50\,\text{L}$ ($\approx 9\,\text{kg}$ capacity is too small; specify the
next standard size up, a $20$ to $25\,\text{kg}$ cylinder) cylinder fits.
The pressure rating is set by the saturation pressure at the highest
foreseeable temperature: at $50\,\degC$ ($T_r = 0.87$) propane's
saturation pressure is about $1.7\,\text{MPa}$ absolute, so a tank rated
to $2.5\,\text{MPa}$ carries margin. State the design temperature
explicitly, because the rating follows the saturation curve.
\end{exercise}

\begin{exercise}
Design the suction-side piping for a centrifugal pump drawing water
at $80\,\degC$ from a tank below the pump (suction lift). State the
maximum allowable suction lift to satisfy the manufacturer's
$\text{NPSH}_r = 3\,\text{m}$, accounting for vapour pressure and
friction losses.

\paragraph{Model-answer sketch.} Required: $\text{NPSH}_a \ge
\text{NPSH}_r = 3\,\text{m}$, with
$\text{NPSH}_a = (p_{\text{atm}} - p_{\text{sat}})/(\rho g) - z_{\text{lift}}
- h_{\text{friction}}$ (lift is negative static head). At $80\,\degC$ the
saturation pressure of water is about $47\,\text{kPa}$, giving an
atmospheric-minus-vapour head of $(101.3 - 47)\times 10^3/(1000\times
9.81) = 5.5\,\text{m}$. Allow $0.5\,\text{m}$ of suction-line friction.
Then $z_{\text{lift}} \le 5.5 - 0.5 - 3 = 2.0\,\text{m}$. The hot water
caps the lift at about $2\,\text{m}$; cold water at $20\,\degC$
($p_{\text{sat}} = 2.3\,\text{kPa}$) would allow nearly $6\,\text{m}$,
which is why feedwater temperature drives suction-side layout.
\end{exercise}

\begin{exercise}
Specify the protection scheme (slow-closing valve, air chamber, or
surge tank) for a $500\,\text{m}$ water-main with maximum flow
velocity $2\,\text{m}/\text{s}$ and rated working pressure
$1.5\,\text{MPa}$. State the closure time required to keep the
water-hammer surge below the rated pressure.

\paragraph{Model-answer sketch.} The Joukowsky rise for instantaneous
closure is $\Delta p = \rho c \Delta v = (1000)(1200)(2) = 2.4\,\text{MPa}$,
above the $1.5\,\text{MPa}$ rating, so instantaneous closure is
forbidden. The round-trip time is $2L/c = 2(500)/1200 = 0.83\,\text{s}$.
For a slow closure the surge scales as $\Delta p \approx \rho c \Delta v
\cdot (2L/c)/t_{\text{close}}$. Setting that below the working margin,
say $1.2\,\text{MPa}$ (allowing for steady working pressure), requires
$t_{\text{close}} \ge (2.4/1.2)(0.83) = 1.7\,\text{s}$. Specify a
slow-closing valve with at least a $2\,\text{s}$ stroke as the primary
control, backed by an air chamber or surge tank sized to absorb the
residual transient. State the closure time on the valve specification.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A centrifugal pump installed in a hot-water service is observed to
produce reduced flow and a characteristic "gravel-rattling" noise
when the suction tank is drawn down to half its level. Identify the
mechanism, the diagnostic measurement, and the corrective action.

\paragraph{Solution sketch.} The mechanism is cavitation: drawing the
tank down lowers the static suction head, so $\text{NPSH}_a$ falls below
$\text{NPSH}_r$ and vapour bubbles form and collapse in the impeller. The
gravel-rattling sound is the broadband collapse noise, and the reduced
flow is the head loss from vapour in the impeller passage. Diagnostic
measurement: the suction-side pressure (a gauge at the pump inlet) read
against the water's vapour pressure at the operating temperature; compute
$\text{NPSH}_a$ and compare to the manufacturer's $\text{NPSH}_r$.
Corrective action: raise $\text{NPSH}_a$ by keeping the tank level higher
(refill threshold), lowering the pump, cooling the water, reducing
suction-line losses, or reducing the flow rate, which lowers
$\text{NPSH}_r$.
\end{exercise}

\begin{exercise}
A residential refrigerator's evaporator coil partially frosts over
within a few minutes of the door being opened, then thaws when the
compressor cycles off. Identify the operating regime on the
refrigerant's $p$-$h$ diagram and the modification (sub-cooling,
defrost timer, expansion-valve adjustment) that would address the
behaviour.

\paragraph{Solution sketch.} Opening the door dumps warm humid air onto
the cold evaporator. The evaporator operates at the low-pressure leg of
the $p$-$h$ loop (states~4 to~1 in
figure~\ref{fig:vol04ch02:ph-refrigerant-cycle}), at a coil temperature
below $0\,\degC$, so water vapour from the room condenses and freezes on
the coil; when the compressor cycles off, the coil warms above freezing
and the frost thaws. The behaviour is the normal frost-thaw cycle, not a
refrigerant fault. The standard modification is a defrost timer (periodic
heater or off-cycle defrost), not an expansion-valve change; reducing
door-open time and the humidity load is the operational fix. Persistent
heavy frosting that does not clear points instead to a low evaporator
temperature from an over-throttled expansion device or low charge.
\end{exercise}

\begin{exercise}
A liquefied-natural-gas tanker truck arrives at a remote storage
facility with the tank's working pressure at the upper alarm
threshold. The truck has been in transit for $36\,\text{h}$ at
ambient $30\,\degC$. Identify whether the pressure rise is
consistent with heat-leak vapourisation and the operational decision
to be made before the cargo is offloaded.

\paragraph{Solution sketch.} Liquefied natural gas is stored as a boiling
cryogen near $-160\,\degC$; heat leaking through the insulation boils off
a small fraction of the cargo (boil-off gas), and in a closed tank that
vapour raises the pressure along the saturation curve. Over $36\,\text{h}$
at $30\,\degC$ ambient a realistic boil-off of a fraction of a percent per
day is fully consistent with a slow climb to the upper alarm threshold; no
leak or fault need be inferred. The operational decision before offloading
is to relieve the accumulated pressure safely, by venting or recovering
boil-off gas to a flare or reliquefaction unit, or by drawing vapour into
the receiving system, so that the tank is brought below the alarm before
liquid transfer begins. Confirm the rise is heat-leak boil-off (gradual,
matching the insulation heat-leak rate) rather than a contamination or
roll-over event, which would show a faster or stepped rise.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A water-supply distribution main ruptures during a routine valve
closure operation. The downstream pressure at the time of failure is
reported as $4\,\text{MPa}$ in a pipe rated for $1.6\,\text{MPa}$.
Compute the Joukowsky pressure rise from the upstream flow velocity
and pipe geometry, and identify the operational discipline that
should have prevented the rupture.

\paragraph{Solution sketch.} A reported transient of $4\,\text{MPa}$ over
a normal working pressure of order $0.5$ to $1\,\text{MPa}$ implies a
Joukowsky surge of roughly $3$ to $3.5\,\text{MPa}$. Reading
$\Delta p = \rho c \Delta v$ backward with $\rho = 1000$ and $c \approx
1200\,\text{m}/\text{s}$ gives $\Delta v = \Delta p/(\rho c) =
3.0\times 10^6/(1.2\times 10^6) = 2.5\,\text{m}/\text{s}$, an entirely
ordinary main velocity stopped too fast. The closure was rapid relative
to the pipe's $2L/c$, so the full Joukowsky rise developed and exceeded
the $1.6\,\text{MPa}$ rating. The operational discipline that should have
prevented it is enforced slow valve closure (stroke time longer than
$2L/c$), backed by surge protection; the rupture is a procedure failure,
not a material defect.
\end{exercise}

\begin{exercise}
A propane storage tank engulfed in fire ruptures after $20\,\text{min}$
of fire exposure. The tank had a thermal-relief valve that did not
prevent the failure. Identify the BLEVE mechanism, the mode of metal
failure (vapour-space wall versus liquid-wetted wall), and the
sprinkler or insulation requirement that would have provided
margin.

\paragraph{Solution sketch.} The mechanism is a BLEVE: external fire
heats the tank, the liquid inside is driven above its atmospheric boiling
point, and when the wall fails the liquid flashes explosively. The metal
fails at the vapour-space wall, not the liquid-wetted wall: the liquid
inside removes heat from the wetted wall by boiling and keeps it near the
liquid temperature, while the vapour-space wall receives the same fire
flux with little internal cooling, so it overheats, loses strength, and
tears first. A thermal-relief valve protects against overpressure but not
against wall weakening, which is why it did not prevent the failure. The
margin comes from cooling or insulating the vapour-space wall: a
water-deluge sprinkler that wets the upper shell, or fire-rated
insulation, keeps the metal below its strength-loss temperature for the
duration of the fire, the controls named in the failure section.
\end{exercise}

\begin{exercise}
A vacuum-condenser circulating-water pump on a power-plant cooling
tower has suffered cavitation pitting on its impeller, requiring
replacement every six months. Identify the cycle's design parameter
that most directly drives the cavitation and the modification that
would extend impeller life.

\paragraph{Solution sketch.} The pump draws warm water from the
condenser hotwell, which sits under the turbine's vacuum, so the water is
close to its saturation pressure at the operating temperature. The design
parameter that most directly drives cavitation is the available NPSH at
the pump suction, set by the hotwell level above the pump minus the
vapour-pressure head; the vacuum condenser leaves little margin. The
modification that extends impeller life is to raise $\text{NPSH}_a$:
increase the static head by raising the hotwell or lowering the pump,
sub-cool the condensate a few degrees below saturation to drop its vapour
pressure, or fit a lower-$\text{NPSH}_r$ impeller (inducer). Cavitation-
resistant materials (stainless, bronze) slow the damage but treat the
symptom rather than the cause.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer proposes to treat steam at $200\,\degC$ and
$1\,\text{MPa}$ as an ideal gas in a turbine-inlet calculation.
Write a one-paragraph reply that names the relevant reduced-property
test and identifies the approximate error.

\paragraph{Solution sketch.} Apply the reduced-property test with water's
critical constants ($T_c = 647\,\si{\kelvin}$, $p_c = 22.06\,\text{MPa}$):
$T_r = 473/647 = 0.73$ and $p_r = 1/22.06 = 0.045$. The low $p_r$ helps,
but $T_r$ is well below the $T_r > 2$ safe band, and steam at
$1\,\text{MPa}$ near saturation is a polar vapour close to its
saturation curve, exactly where the ideal-gas law fails. The
compressibility factor here is roughly $0.93$ to $0.95$, so treating the
steam as ideal misstates the specific volume and enthalpy by several
percent, enough to matter for a turbine-inlet calculation. The
disciplined move is to read the steam tables, not the ideal-gas law.
\end{exercise}

\begin{exercise}
A facilities manager argues that water-hammer protection is
"unnecessary on plastic pipes because the pipe absorbs the surge."
Write a one-paragraph reply identifying the partial truth in the
claim and the failure modes that remain.

\paragraph{Solution sketch.} The partial truth is real: a plastic pipe
has a lower elastic modulus than steel, so the wave speed $c$ in a
plastic-walled pipe is lower (often $300$ to $500\,\text{m}/\text{s}$
against $1200\,\text{m}/\text{s}$ for steel), and since $\Delta p = \rho c
\Delta v$ the Joukowsky surge is proportionally smaller. The claim fails
because the surge is reduced, not removed: a $400\,\text{m}/\text{s}$
plastic main carrying water at $2\,\text{m}/\text{s}$ still sees
$\Delta p = (1000)(400)(2) = 0.8\,\text{MPa}$, comparable to working
pressure. Plastic pipe also creeps and fatigues under repeated pressure
cycles, and its fittings and joints remain vulnerable. Slow valve
closure and surge protection remain required; the lower wave speed buys
margin, it does not grant immunity.
\end{exercise}
