\chapter{Eigenvalues and decompositions}
\label{vol02:ch10}

% \spn, \col, \nul, \rank, \sgn, \proj, \tr, \diag, \argmin, \argmax
% are defined in eng-macros.sty.

\epigraph{Pick the right basis and the matrix becomes a list of
numbers. Pick the wrong basis and it stays an opaque grid.}{Working
paraphrase of the spectral programme of Strang,
\emph{Introduction to Linear Algebra}, 5th ed.\
\cite[ch.~6]{text:strang2016}.}

\chapmeta{Half-life: foundational. Archetypes: scaling (eigenvalues
as the scalar action of a matrix on its preferred directions),
control (eigenvalue location as the test of linear stability).
Exercise target: 35. Project track: simulation with analysis
(implement Principal Component Analysis from scratch on a real
dataset, with a 1500-word reflection). Hazard class: none. Reader
path default: standard, with sections explicitly tagged. Named
cases: none required; the failure section treats defective and
complex-eigenvalue regimes generically.}

\noindent
Chapter 9 read a matrix as a linear map between two coordinate
spaces and asked how to solve, project, and condition it. The next
question is internal. Given a square matrix, are there directions
the matrix leaves alone, scaling them but not rotating them? When
such directions exist and span the input space, the matrix is a
diagonal scaling in disguise. When they do not, the matrix has
structure that diagonal scaling cannot describe, and the engineer
needs a different decomposition. The singular value decomposition
covers every matrix without exception and is the workhorse of this
chapter and the rest of the volume.

\begin{mastery}
\textit{Historical sketch: where the word eigenvalue comes from.}
The idea predates the name by more than a century. Euler and Lagrange
met eigenvalues in the eighteenth century while studying the rotation
of rigid bodies and the small oscillations of coupled pendulums; the
``principal axes'' of an inertia tensor and the ``normal modes'' of a
vibrating system are eigenvectors, computed long before the modern
vocabulary existed. Cauchy, in the 1820s, proved that the eigenvalues
of a symmetric matrix (in his language, the roots of the
characteristic equation of a quadratic form) are real, the result
this chapter calls part of the spectral theorem. The German word
\emph{eigen} means ``own'' or ``characteristic,'' and Hilbert used
\emph{Eigenwert} (own-value) in his work on integral equations around
1904; the hybrid English ``eigenvalue'' settled into use in the 1920s
and 1930s, pushed by the quantum-mechanics literature, where the
allowed energies of a system are eigenvalues of an operator. The
singular value decomposition has a parallel lineage: Beltrami (1873)
and Jordan (1874) found it independently for real square matrices,
Eckart and Young (1936) proved its optimality for low-rank
approximation, and Golub and collaborators turned it into a stable
numerical algorithm in the 1960s. The reader meets, in one chapter,
two hundred years of accumulated apparatus. The vocabulary is recent;
the mathematics is old, and it was driven from the start by physical
problems of vibration, rotation, and stability.
\end{mastery}

\section{Eigenvalues and eigenvectors}\pathtag{core}

A square matrix $\mat{A} \in \R^{n \times n}$ acts on a vector
$\vect{x} \in \R^{n}$ by $\vect{x} \mapsto \mat{A}\vect{x}$. For most
$\vect{x}$, the output points in a different direction from the
input. For a privileged few, the matrix only scales: $\mat{A}
\vect{x} = \lambda \vect{x}$ for some scalar $\lambda$. These
privileged directions are the matrix's eigenvectors, and the scaling
factors are its eigenvalues.

\subsection{Definition and first identification}

An \engterm{eigenvector} of $\mat{A}$ is a nonzero vector $\vect{x}
\in \C^{n}$ satisfying $\mat{A}\vect{x} = \lambda \vect{x}$ for
some scalar $\lambda \in \C$. The scalar $\lambda$ is the
corresponding \engterm{eigenvalue}. We work over $\C$ because a real
matrix may have complex eigenvalues even when every entry is real,
as section 10.7 makes precise. The eigenvalue equation rearranges to

\[
(\mat{A} - \lambda \mat{I}) \vect{x} = \vect{0},
\]

which has a nonzero solution if and only if $\mat{A} - \lambda
\mat{I}$ is singular, that is, if and only if its determinant
vanishes. The \engterm{characteristic polynomial} of $\mat{A}$ is

\[
p_{\mat{A}}(\lambda) = \det(\mat{A} - \lambda \mat{I}),
\]

a polynomial of degree $n$ in $\lambda$ with leading coefficient
$(-1)^{n}$. Its roots are the eigenvalues of $\mat{A}$. The
fundamental theorem of algebra guarantees $n$ roots in $\C$, counted
with multiplicity. A real matrix has eigenvalues that are real, or
that come in complex conjugate pairs.

For a $2 \times 2$ matrix the characteristic polynomial is

\[
p_{\mat{A}}(\lambda) = \lambda^{2} - (\tr \mat{A}) \lambda
+ \det \mat{A},
\]

with $\tr \mat{A} = A_{11} + A_{22}$ the sum of diagonal entries.
The trace is the sum of eigenvalues; the determinant is their
product. Both relations generalise: for any $n \times n$ matrix,
$\tr \mat{A} = \sum_{i} \lambda_{i}$ and $\det \mat{A} = \prod_{i}
\lambda_{i}$. The two invariants summarise the spectrum from
quantities that are cheap to compute directly from the entries.

\begin{mastery}
The trace and determinant are the first and last of a full family. The
characteristic polynomial of an $n \times n$ matrix,
\[
p_{\mat{A}}(\lambda) = (-1)^{n}\bigl(\lambda^{n} - c_{1}\lambda^{n-1}
+ c_{2}\lambda^{n-2} - \cdots + (-1)^{n}c_{n}\bigr),
\]
has coefficients $c_{k}$ that are the \engterm{elementary symmetric
functions} of the eigenvalues: $c_{1} = \sum_i \lambda_i = \tr\mat{A}$,
$c_{2} = \sum_{i<j}\lambda_i\lambda_j$, up to $c_{n} = \prod_i
\lambda_i = \det\mat{A}$. Each $c_{k}$ also equals the sum of the
$k \times k$ principal minors of $\mat{A}$, so all $n$ invariants are
computable from the entries without ever finding an eigenvalue. The
middle coefficients are not idle. For a $3 \times 3$ matrix, $c_{2} =
\tfrac{1}{2}\bigl((\tr\mat{A})^{2} - \tr(\mat{A}^{2})\bigr)$ is the
second invariant of a stress or inertia tensor, the quantity that,
with the trace and determinant, forms the three coordinate-free
numbers a continuum-mechanics analysis reports. The power sums
$p_k = \sum_i \lambda_i^k = \tr(\mat{A}^k)$ connect to the $c_k$
through Newton's identities, which is how a spectral quantity such as
$\sum_i \lambda_i^{2}$ is read off $\tr(\mat{A}^2)$ directly rather
than by squaring each eigenvalue. Every symmetric function of the
spectrum is a function of these invariants, and the invariants are
functions of the entries; the eigenvalues sit in between as the roots
that tie the two descriptions together.
\end{mastery}

\subsection{Geometric and algebraic multiplicity}

An eigenvalue $\lambda$ is a root of the characteristic polynomial.
Its \engterm{algebraic multiplicity} is the multiplicity of that
root, the exponent of $(\lambda_{0} - \lambda)$ in the factorisation
of $p_{\mat{A}}$. The set of vectors satisfying $(\mat{A} -
\lambda_{0} \mat{I}) \vect{x} = \vect{0}$ is the \engterm{eigenspace}
of $\lambda_{0}$, written $E_{\lambda_{0}} = \nul(\mat{A} -
\lambda_{0} \mat{I})$. Its dimension is the \engterm{geometric
multiplicity} of $\lambda_{0}$.

The geometric multiplicity is at most the algebraic multiplicity. A
$3 \times 3$ matrix with characteristic polynomial $(\lambda - 2)^{3}$
has algebraic multiplicity three at $\lambda = 2$. Its geometric
multiplicity at $2$ is one, two, or three. When it is three, the
matrix is the diagonal $2 \mat{I}$ in some basis. When it is one,
the matrix has only a one-dimensional eigenspace and is
\engterm{defective}: it cannot be diagonalised. The defective case
is where the engineer's panic in section 10.7 begins.

\subsection{Similarity and what the spectrum does not depend on}

Two matrices $\mat{A}$ and $\mat{B}$ are \engterm{similar} when
$\mat{B} = \mat{S}\inv \mat{A} \mat{S}$ for some invertible
$\mat{S}$. Similarity is the algebraic statement that $\mat{A}$ and
$\mat{B}$ are the same linear map written in two different bases.
Similar matrices share their characteristic polynomial, hence their
eigenvalues (with multiplicities), their trace, and their
determinant. The proof is one line: $\det(\mat{S}\inv \mat{A}
\mat{S} - \lambda \mat{I}) = \det(\mat{S}\inv (\mat{A} - \lambda
\mat{I}) \mat{S}) = \det(\mat{S}\inv) \det(\mat{A} - \lambda
\mat{I}) \det(\mat{S}) = \det(\mat{A} - \lambda \mat{I})$, since the
determinant of $\mat{S}\inv$ cancels the determinant of $\mat{S}$.

This is why the eigenvalues deserve to be called the matrix's
intrinsic numbers. Every quantity the engineer reads off a square
matrix that should not depend on the choice of coordinates, the
determinant, the trace, the rank, the spectral radius, the
conditioning, is a function of the spectrum and is preserved under
similarity. Diagonalisation, when it exists, is the choice of basis
that makes the map maximally transparent: $\mat{A} = \mat{S}
\mat{\Lambda} \mat{S}\inv$ says $\mat{A}$ is similar to the diagonal
matrix of its eigenvalues. The whole programme of the chapter is to
find the basis in which a given matrix is as close to diagonal as it
can be made.

\subsection{The Rayleigh quotient}

For a symmetric matrix $\mat{A}$ and a nonzero vector $\vect{x}$,
the \engterm{Rayleigh quotient} is

\[
R(\vect{x}) = \frac{\vect{x}\transpose \mat{A} \vect{x}}
{\vect{x}\transpose \vect{x}}.
\]

When $\vect{x}$ is an eigenvector $\vect{v}_{i}$, the quotient
returns the eigenvalue: $R(\vect{v}_{i}) = \lambda_{i}$, since
$\mat{A} \vect{v}_{i} = \lambda_{i} \vect{v}_{i}$ collapses the
numerator to $\lambda_{i} \vect{v}_{i}\transpose \vect{v}_{i}$. For
a general $\vect{x}$ the quotient lies between the smallest and
largest eigenvalues, $\lambda_{\min} \leq R(\vect{x}) \leq
\lambda_{\max}$, and the extremes are achieved exactly at the
corresponding eigenvectors. This is the variational characterisation
of eigenvalues, the engine behind the power method's eigenvalue
estimate (section 10.6) and behind the Courant-Fischer minimax
theorem that proves the spectral theorem and the Eckart-Young
theorem.

The Rayleigh quotient is the engineer's cheap eigenvalue estimate.
Given an approximate eigenvector $\vect{x}$, $R(\vect{x})$ is a far
better estimate of the eigenvalue than $\vect{x}$ is of the
eigenvector: the error in $R(\vect{x})$ is of order the square of
the error in $\vect{x}$, because $R$ is stationary at the
eigenvectors. A vector that is $1\%$ off the true eigenvector
direction gives a Rayleigh quotient that is about $0.01\%$ off the
true eigenvalue. The quadratic accuracy is why production
eigensolvers report eigenvalues by Rayleigh quotients of their
iterates rather than by any cruder formula.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-rayleigh-quotient.tex}

\Cref{fig:vol02:ch10:rayleigh-quotient} makes the trapping visible for
the symmetric matrix $\begin{psmallmatrix} 4 & 1 \\ 1 & 4
\end{psmallmatrix}$. As the input unit vector sweeps through every
direction, the quotient traces a curve that never leaves the band
between $\lambda_{\min} = 3$ and $\lambda_{\max} = 5$ and touches each
extreme exactly at the corresponding eigenvector. The flatness of the
curve at each touch point is the stationarity that gives the quadratic
accuracy; away from the touch points the quotient moves at full speed,
so a rough eigenvector guess still lands a good eigenvalue estimate.

\begin{example}
The quadratic accuracy is worth watching on numbers. Take the same
matrix $\mat{A} = \begin{psmallmatrix} 4 & 1 \\ 1 & 4
\end{psmallmatrix}$, whose top eigenvector is $\vect{v} = (1, 1)/
\sqrt{2}$ at eigenvalue $\lambda = 5$. Feed in a vector $6^{\circ}$
off that direction, $\vect{x} = (\cos 51^{\circ}, \sin 51^{\circ}) =
(0.629, 0.777)$. The eigenvector error is the sine of the offset
angle, $\sin 6^{\circ} \approx 0.105$, about $10\%$. The Rayleigh
quotient is $R(\vect{x}) = \vect{x}\transpose\mat{A}\vect{x} =
4 + 2(0.629)(0.777) = 4 + 0.977 = 4.977$, an eigenvalue error of
$0.023$, or $0.46\%$ of the eigenvalue. A $10\%$ error in the vector
has become a sub-half-percent error in the eigenvalue, the square-law
gain $R$ delivers because it is stationary at $\vect{v}$. The same
input reports the eigenvalue two orders of magnitude more accurately
than it reports the eigenvector, which is why an eigensolver quotes
$R$ of its current iterate as its running eigenvalue estimate.
\end{example}

\subsection{Gerschgorin discs: bounding the spectrum from the entries}

The eigenvalues of a matrix can be localised, without computing
them, directly from its entries. For each row $i$ of $\mat{A}$,
define the \engterm{Gerschgorin disc} as the closed disc in the
complex plane centred at the diagonal entry $A_{ii}$ with radius
$r_{i} = \sum_{j \neq i} \abs{A_{ij}}$, the sum of the
off-diagonal magnitudes in that row.

\begin{principle}[Gerschgorin circle theorem]
Every eigenvalue of $\mat{A}$ lies in at least one of the
Gerschgorin discs. If a union of $k$ discs is disjoint from the
remaining $n - k$, that union contains exactly $k$ eigenvalues,
counted with multiplicity.
\end{principle}

The proof is short. Let $\lambda$ be an eigenvalue with eigenvector
$\vect{x}$, and let $i$ be the index of the largest-magnitude entry
of $\vect{x}$, so $\abs{x_{i}} \geq \abs{x_{j}}$ for all $j$. The
$i$-th row of $\mat{A} \vect{x} = \lambda \vect{x}$ reads
$\sum_{j} A_{ij} x_{j} = \lambda x_{i}$, which rearranges to
$(\lambda - A_{ii}) x_{i} = \sum_{j \neq i} A_{ij} x_{j}$. Taking
magnitudes and dividing by $\abs{x_{i}}$,
$\abs{\lambda - A_{ii}} \leq \sum_{j \neq i} \abs{A_{ij}}
\abs{x_{j}}/\abs{x_{i}} \leq \sum_{j \neq i} \abs{A_{ij}} = r_{i}$,
since each ratio $\abs{x_{j}}/\abs{x_{i}}$ is at most one. So
$\lambda$ lies in the $i$-th disc.

Gerschgorin is the engineer's first, free diagnostic. A strictly
diagonally dominant matrix (each $\abs{A_{ii}} > r_{i}$) has every
disc clear of the origin, so no disc contains zero, so no eigenvalue
is zero, so the matrix is invertible, all established without a
single floating-point operation on the matrix itself. The stability
bound for stochastic matrices in section 10.5 is a Gerschgorin
argument; so is the claim that the eigenvalues of a perturbed
diagonal matrix stay near the diagonal entries when the
perturbation is small.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-gerschgorin-discs.tex}

\Cref{fig:vol02:ch10:gerschgorin-discs} draws the discs of a
representative $4 \times 4$ matrix together with its true spectrum. An
isolated disc, disjoint from the rest, holds exactly the number of
eigenvalues equal to the count of centres it merges, by the second
half of the theorem; the engineer reads off, from the entries alone,
that two eigenvalues live near the well-separated diagonal entries and
two share a merged region. The disc radii also bound how far an
eigenvalue can move under a perturbation of the off-diagonal entries,
which is the next subsection's subject.

\begin{example}
The counting half of the theorem, the part that says a cluster of $k$
discs holds exactly $k$ eigenvalues, is where Gerschgorin earns real
work. Take
\[
\mat{A} = \begin{pmatrix} 1 & 0.1 & 0.1 \\ 0.2 & 5 & 0.3 \\
0.1 & 0.2 & 9 \end{pmatrix}.
\]
The three discs are centred at $1$, $5$, $9$ with radii $r_1 = 0.2$,
$r_2 = 0.5$, $r_3 = 0.3$, giving intervals $[0.8, 1.2]$, $[4.5, 5.5]$,
$[8.7, 9.3]$ on the real axis. The three intervals are pairwise
disjoint, so each isolated disc contains exactly one eigenvalue by the
second half of the theorem: one eigenvalue near $1$, one near $5$, one
near $9$, each pinned to a narrow band without solving a cubic. The
true eigenvalues, computed for the check, are $0.994$, $4.99$, and
$9.02$, each comfortably inside its own disc and near the disc centre
because the off-diagonal entries are small. The engineer reads off,
from the entries alone, that this matrix has three well-separated real
eigenvalues and that a shift-and-invert solver targeting any one of
them needs only to place its shift at the corresponding diagonal entry.
Widen the off-diagonals until two discs overlap, say by raising the
$(2,3)$ and $(3,2)$ entries so that $r_2$ and $r_3$ grow past $2$, and
the union of those two discs then holds two eigenvalues that
Gerschgorin can no longer separate individually; the count survives the
merge, but the isolation does not.
\end{example}

\subsection{Eigenvalue perturbation: how far a spectrum can move}\pathtag{mastery}

The engineer rarely has the exact matrix. Entries carry measurement
noise, model error, and floating-point rounding, and the question is
how much the spectrum moves when the matrix moves. For symmetric
matrices the answer is reassuring and sharp.

\begin{principle}[Weyl's inequality]
Let $\mat{A}$ and $\mat{A} + \mat{E}$ be symmetric $n \times n$
matrices with eigenvalues $\lambda_{1} \geq \cdots \geq \lambda_{n}$
and $\mu_{1} \geq \cdots \geq \mu_{n}$ respectively, both ordered
descending. Then $\abs{\lambda_{i} - \mu_{i}} \leq
\norm{\mat{E}}_{2}$ for every $i$.
\end{principle}

A symmetric matrix is perfectly conditioned for its eigenvalues: a
perturbation of spectral-norm size $\epsilon$ moves every eigenvalue
by at most $\epsilon$. A stiffness matrix whose entries are known to
$1\%$ has natural frequencies known to comparable relative accuracy,
and the engineer can quote a frequency with an honest error bar
without recomputing the spectrum for every entry within tolerance.

The non-symmetric case is the one that ambushes the unwary. The
relevant statement is the Bauer-Fike theorem: for a diagonalisable
$\mat{A} = \mat{S} \mat{\Lambda} \mat{S}\inv$ and any eigenvalue
$\mu$ of $\mat{A} + \mat{E}$, there is an eigenvalue $\lambda_{i}$ of
$\mat{A}$ with $\abs{\mu - \lambda_{i}} \leq \kappa_{2}(\mat{S})
\norm{\mat{E}}_{2}$. The eigenvalue condition number is
$\kappa_{2}(\mat{S})$, the conditioning of the eigenvector matrix,
not of $\mat{A}$ itself. A non-normal matrix with nearly parallel
eigenvectors has $\kappa_{2}(\mat{S})$ large, and a tiny perturbation
can move its eigenvalues by a large multiple of $\norm{\mat{E}}$.
The classic demonstration is the $n \times n$ Jordan-like matrix with
$\lambda$ on the diagonal and $1$ on the superdiagonal: perturbing
the bottom-left corner entry from $0$ to $\epsilon$ splits the single
eigenvalue $\lambda$ into $n$ values spread on a circle of radius
$\epsilon^{1/n}$ about $\lambda$. At $n = 10$ and $\epsilon =
10^{-10}$, the radius is $10^{-1} = 0.1$: a perturbation at the level
of double-precision rounding moves the eigenvalues by a tenth. The
spectrum of a strongly non-normal matrix is a number the engineer
should distrust, and the pseudospectrum of section 10.7 is the
honest replacement.

\subsection{Worked example: a $3 \times 3$ matrix end to end}

The two-by-two case hides the bookkeeping that makes larger
problems realistic. Consider

\[
\mat{A} = \begin{pmatrix} 2 & 0 & 0 \\ 1 & 2 & 0 \\ 1 & 1 & 3
\end{pmatrix},
\]

a lower-triangular matrix. The eigenvalues of a triangular matrix
are its diagonal entries, because $\det(\mat{A} - \lambda \mat{I})$
expands to the product of the diagonal entries of the triangular
matrix $\mat{A} - \lambda \mat{I}$: here
$(2 - \lambda)(2 - \lambda)(3 - \lambda)$. The eigenvalues are
$\lambda = 2$ (algebraic multiplicity two) and $\lambda = 3$
(algebraic multiplicity one). Gerschgorin offers an independent
sanity check: the discs are centred at $2$, $2$, $3$ with radii
$0$, $1$, $2$, and the eigenvalues $2, 2, 3$ all fall inside the
union $[1, 5]$ on the real axis, as required.

The eigenspace at $\lambda = 3$ is the null space of

\[
\mat{A} - 3\mat{I} = \begin{pmatrix} -1 & 0 & 0 \\ 1 & -1 & 0 \\
1 & 1 & 0 \end{pmatrix}.
\]

Row reduction forces $x_{1} = 0$ from the first row, then
$x_{1} - x_{2} = 0$ from the second gives $x_{2} = 0$, leaving
$x_{3}$ free. The eigenvector is $\vect{v}_{3} = (0, 0, 1)
\transpose$.

The eigenspace at $\lambda = 2$ is the null space of

\[
\mat{A} - 2\mat{I} = \begin{pmatrix} 0 & 0 & 0 \\ 1 & 0 & 0 \\
1 & 1 & 1 \end{pmatrix}.
\]

The second row forces $x_{1} = 0$; the third row then forces
$x_{2} + x_{3} = 0$, that is $x_{3} = -x_{2}$. The eigenspace is
one-dimensional, spanned by $\vect{v}_{2} = (0, 1, -1)\transpose$.
Algebraic multiplicity two, geometric multiplicity one: the matrix
is defective. There are only two independent eigenvectors,
$\vect{v}_{2}$ and $\vect{v}_{3}$, for a three-dimensional space.
The matrix cannot be diagonalised, and the apparatus of section 10.7
applies. The trace check still holds, $2 + 2 + 3 = 7 = \tr \mat{A}$,
because the trace counts eigenvalues with algebraic multiplicity
regardless of defectiveness; the determinant check gives
$2 \cdot 2 \cdot 3 = 12 = \det \mat{A}$. The two invariants are
blind to the defect, which is one reason the defect is easy to miss.

\subsection{Computation by hand for small matrices}

For $n \leq 4$, the characteristic polynomial is short and the
roots can be found by inspection or by the quadratic, cubic, or
quartic formula. Beyond $n = 4$ the polynomial-roots route is
useless. The Abel-Ruffini theorem rules out closed-form solution by
radicals for general degree-five and higher polynomials, and even
for $n = 4$ the formulas are numerically poor. Production
eigensolvers do not factor the characteristic polynomial. They
apply iterative methods (QR algorithm with shifts) directly to
$\mat{A}$ and read off eigenvalues from the limit of the iteration
\cite[ch.~25]{text:trefethen-bau1997}. The reader works small cases
by hand to internalise the structure; for any $n$ above about $4$
the reader uses a library routine.

\begin{estimation}
\textbf{Question.} A spring-mass system has two equal masses $m$
connected to walls and to each other by three identical springs of
stiffness $k$, in a line. The equations of motion are
$m \ddot{x}_{i} = $ net spring force on mass $i$. The natural
frequencies of vibration are $\omega = \sqrt{\lambda}$, where
$\lambda$ are eigenvalues of $\mat{K}/m$ for the stiffness matrix
$\mat{K}$. Estimate the ratio of the highest to the lowest natural
frequency.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The two physical modes are: the masses moving in phase (slow,
because the centre spring does not stretch), and the masses moving
out of phase (fast, because the centre spring stretches double). In
phase, each mass sees only its wall spring; the restoring stiffness
is $k$, and $\omega_{1}^{2} = k/m$. Out of phase, each mass sees
its wall spring plus twice the centre spring; the restoring stiffness
is $3k$, and $\omega_{2}^{2} = 3k/m$. The ratio is
$\omega_{2}/\omega_{1} = \sqrt{3} \approx 1.7$. The spectral
calculation in section 10.5's exercises confirms the estimate.
\end{estimation}

The estimation reads the modes off the geometry; the algebra
confirms them. This is the order in which the engineer should learn
to think. Modes first, eigenvalues second. The eigenvalues attach
numbers to a structure the geometry already names.

\subsection{Worked example: a $2 \times 2$ matrix}

Consider

\[
\mat{A} = \begin{pmatrix} 4 & 1 \\ 2 & 3 \end{pmatrix}.
\]

The trace is $7$ and the determinant is $4 \cdot 3 - 1 \cdot 2 =
10$. The characteristic polynomial is $\lambda^{2} - 7\lambda + 10 =
(\lambda - 5)(\lambda - 2)$. The eigenvalues are $\lambda_{1} = 5$
and $\lambda_{2} = 2$, both real and distinct.

The eigenspace at $\lambda_{1} = 5$ comes from the null space of

\[
\mat{A} - 5 \mat{I} = \begin{pmatrix} -1 & 1 \\ 2 & -2 \end{pmatrix},
\]

whose rows are dependent. The condition $-x_{1} + x_{2} = 0$ gives
$\vect{v}_{1} = (1, 1)\transpose$ as a basis for $E_{5}$. Similarly,
the eigenspace at $\lambda_{2} = 2$ comes from the null space of

\[
\mat{A} - 2 \mat{I} = \begin{pmatrix} 2 & 1 \\ 2 & 1 \end{pmatrix},
\]

with condition $2 x_{1} + x_{2} = 0$, giving $\vect{v}_{2} = (1, -2)
\transpose$ as a basis for $E_{2}$. The two eigenvectors are
linearly independent. They span $\R^{2}$. The matrix is
diagonalisable, and section 10.2 shows what the diagonalisation
buys.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-eigenvector-action.tex}

\Cref{fig:vol02:ch10:eigenvector-action} shows the geometry directly.
A fan of input vectors of equal length, each at a different angle,
rotates as well as stretches under multiplication by $\mat{A}$;
almost every input direction is mapped to a new direction. The two
red arrows are the eigenvectors $\vect{v}_{1}$ and $\vect{v}_{2}$,
and their images $5\vect{v}_{1}$ and $2\vect{v}_{2}$ point along the
same lines (one with reversed orientation if the eigenvalue is
negative). The diagonalisation of the next section is the algebraic
restatement of this picture: change to the eigenbasis, and the
rotation disappears.

\begin{archetype}[Scaling]
The eigenvalues are the scalings the matrix performs on its
preferred directions. Volume I introduced scaling as the dimensional
discipline of how a quantity changes under change of unit; here the
archetype recurs in matrix form. A linear map with eigenvalues
$\lambda_{1}, \ldots, \lambda_{n}$ is, in the right basis, the
diagonal scaling $\diag(\lambda_{1}, \ldots, \lambda_{n})$. Every
question the engineer asks about a square matrix that does not
depend on basis, the determinant, the trace, the spectral norm,
the conditioning, the stability, is a question about its spectrum.
\end{archetype}

\subsection{Worked example: a $3 \times 3$ with a complex pair}

A real matrix can carry a rotation, and rotations have no real
eigenvectors. The example to keep is

\[
\mat{B} = \begin{pmatrix} 3 & 1 & 0 \\ -1 & 3 & 0 \\ 0 & 0 & 5
\end{pmatrix}.
\]

The block structure splits the work. The bottom-right $1 \times 1$
block contributes the real eigenvalue $\lambda_{3} = 5$ with
eigenvector $\vect{v}_{3} = (0, 0, 1)\transpose$. The top-left
$2 \times 2$ block $\begin{psmallmatrix} 3 & 1 \\ -1 & 3
\end{psmallmatrix}$ has trace $6$ and determinant $9 + 1 = 10$, so
its characteristic polynomial is $\lambda^{2} - 6\lambda + 10$, with
roots $\lambda = 3 \pm i$. The full spectrum is $\{3 + i,\, 3 - i,\,
5\}$. The trace check holds, $(3 + i) + (3 - i) + 5 = 11 = \tr
\mat{B}$, with the imaginary parts cancelling because the complex
eigenvalues come as a conjugate pair; the determinant check holds,
$(3 + i)(3 - i)(5) = 10 \cdot 5 = 50 = \det \mat{B}$, with the
product of the conjugate pair returning the real value $\abs{3 +
i}^{2} = 10$. The two invariants stay real even though two of the
three eigenvalues do not, which is the signature of a real matrix
with a complex pair.

The eigenvector at $3 + i$ is complex. Solving $(\mat{B} - (3 + i)
\mat{I})\vect{x} = \vect{0}$ on the top-left block gives the relation
$-i x_{1} + x_{2} = 0$, so $\vect{w} = (1, i, 0)\transpose$ spans the
complex eigenspace; the eigenvector at $3 - i$ is its conjugate
$\bar{\vect{w}} = (1, -i, 0)\transpose$. The engineer who needs a
real basis does not diagonalise over $\C$. The real and imaginary
parts $\operatorname{Re}\vect{w} = (1, 0, 0)\transpose$ and
$\operatorname{Im}\vect{w} = (0, 1, 0)\transpose$ span a real
two-dimensional invariant subspace on which $\mat{B}$ acts as the
scaled rotation $\begin{psmallmatrix} 3 & 1 \\ -1 & 3
\end{psmallmatrix}$. This invariant-plane reading, rather than a
complex eigenvector, is what the real Schur form of section 10.7
returns and what a vibration or control analysis actually uses: the
complex pair $3 \pm i$ says the mode spirals, with growth or decay
rate set by the real part $3$ and oscillation rate set by the
imaginary part $1$.

\section{Diagonalisation}\pathtag{core}

A square matrix $\mat{A}$ is \engterm{diagonalisable} when there
exist an invertible $\mat{S}$ and a diagonal $\mat{\Lambda}$ such
that $\mat{A} = \mat{S} \mat{\Lambda} \mat{S}\inv$. The columns of
$\mat{S}$ are eigenvectors of $\mat{A}$; the diagonal entries of
$\mat{\Lambda}$ are the corresponding eigenvalues. Diagonalisation
exists exactly when the eigenvectors form a basis of $\R^{n}$ (or
$\C^{n}$).

\subsection{When diagonalisation works}

The simplest sufficient condition: if $\mat{A}$ has $n$ distinct
eigenvalues, it is diagonalisable. Eigenvectors corresponding to
distinct eigenvalues are linearly independent, so $n$ distinct
eigenvalues yield $n$ independent eigenvectors and a basis. The
$2 \times 2$ worked example of section 10.1 had distinct eigenvalues
$5$ and $2$, and its eigenvectors $(1, 1)\transpose$ and $(1, -2)
\transpose$ form an invertible $\mat{S}$.

A second sufficient condition: if $\mat{A}$ is real symmetric, it is
diagonalisable, with real eigenvalues and an orthonormal basis of
eigenvectors. This is the \engterm{spectral theorem} for symmetric
matrices. We quote it; the proof uses the Courant-Fischer
characterisation of eigenvalues and is one page in any text
\cite[ch.~6]{text:strang2016}. The orthonormal eigenbasis is the
fact that powers all of structural dynamics, principal component
analysis, and the spectral methods of section 10.4.

A third sufficient condition: if $\mat{A}$ is normal ($\mat{A}\transpose
\mat{A} = \mat{A}\mat{A}\transpose$ for real, $\mat{A}^{*}\mat{A} =
\mat{A}\mat{A}^{*}$ for complex), it is unitarily diagonalisable.
Symmetric matrices are normal; orthogonal matrices are normal;
skew-symmetric matrices are normal. Many matrices the engineer meets
in practice are normal because of physical symmetry.

\subsection{When diagonalisation fails}

A matrix fails to diagonalise when it has fewer linearly independent
eigenvectors than its size. The canonical example is the Jordan
block

\[
\mat{J} = \begin{pmatrix} \lambda & 1 \\ 0 & \lambda \end{pmatrix}.
\]

The characteristic polynomial is $(\lambda - \lambda)^{2}$, with a
single eigenvalue $\lambda$ of algebraic multiplicity two. The
eigenspace is the null space of $\mat{J} - \lambda \mat{I} =
\begin{pmatrix} 0 & 1 \\ 0 & 0 \end{pmatrix}$, which is one
dimensional, spanned by $(1, 0)\transpose$. The geometric multiplicity
is one. The Jordan block is defective.

The general fact is the \engterm{Jordan canonical form}: every
square matrix is similar to a block-diagonal matrix whose blocks
are Jordan blocks. We quote the result and do not prove it. For
working purposes the engineer should know that defective matrices
exist, that the failure is not pathological but generic in
parameter-dependent settings (a parameter sweep that crosses an
eigenvalue collision typically produces a defective matrix at the
collision point), and that production code uses the Schur
decomposition (section 10.7) rather than Jordan form because the
Schur decomposition is numerically stable and the Jordan form is
not.

\subsection{What diagonalisation buys}

When $\mat{A} = \mat{S} \mat{\Lambda} \mat{S}\inv$, powers of
$\mat{A}$ become diagonal in the eigenbasis. For any positive
integer $k$,

\[
\mat{A}^{k} = \mat{S} \mat{\Lambda}^{k} \mat{S}\inv,
\]

and $\mat{\Lambda}^{k}$ is the diagonal matrix with entries
$\lambda_{i}^{k}$. A computation that would have cost $\oom{n^{3}}$
flops per matrix multiplication, repeated $k$ times for $\oom{k
n^{3}}$ flops, becomes $\oom{n^{3}}$ flops for the diagonalisation
plus $\oom{k n}$ flops for the diagonal powers and $\oom{n^{3}}$
flops for the change of basis. The same identity extends to matrix
functions: $f(\mat{A}) = \mat{S} f(\mat{\Lambda}) \mat{S}\inv$ for
any function $f$ that makes sense on the eigenvalues, where
$f(\mat{\Lambda})$ applies $f$ entrywise on the diagonal. The
matrix exponential $e^{\mat{A} t} = \mat{S} e^{\mat{\Lambda} t}
\mat{S}\inv$ is the workhorse of linear-system response (Volume III
Chapter 7 closed-form ODE solutions; Volume IX Chapter 4
state-space dynamics), and its computation reduces to $n$ scalar
exponentials when $\mat{A}$ is diagonalisable.

\subsection{Worked example: powers and the matrix exponential}

For the $\mat{A}$ of section 10.1 with $\mat{S} = \begin{pmatrix} 1
& 1 \\ 1 & -2 \end{pmatrix}$ and $\mat{\Lambda} = \diag(5, 2)$, the
ten-th power is

\[
\mat{A}^{10} = \mat{S} \diag(5^{10}, 2^{10}) \mat{S}\inv.
\]

The dominant entry is $5^{10} = 9{,}765{,}625$, and $2^{10} = 1024$
is negligible by comparison. The asymptotic behaviour of $\mat{A}^{k}
\vect{x}$ for large $k$ is dictated by the dominant eigenvalue and
its eigenvector. This observation is the seed of the power iteration
algorithm of section 10.6.

\begin{mastery}
A non-diagonalisable matrix admits a Jordan decomposition $\mat{A} =
\mat{S} \mat{J} \mat{S}\inv$, with $\mat{J}$ block diagonal in
Jordan blocks. Powers and exponentials of $\mat{J}$ have closed
forms but include polynomial factors in $k$ alongside the geometric
$\lambda^{k}$. For a $2 \times 2$ Jordan block at eigenvalue
$\lambda$, $\mat{J}^{k} = \begin{pmatrix} \lambda^{k} & k \lambda^{k-1}
\\ 0 & \lambda^{k} \end{pmatrix}$. The polynomial growth in $k$ is
the signature of a defective system; it shows up in physical
dynamics as a transient that grows like $t e^{\lambda t}$ alongside
the pure exponential. We return to it in section 10.7.
\end{mastery}

\subsection{Cayley-Hamilton and the algebra of a matrix}\pathtag{mastery}

A matrix satisfies its own characteristic polynomial. The
\engterm{Cayley-Hamilton theorem} states that if $p_{\mat{A}}
(\lambda) = \det(\mat{A} - \lambda \mat{I})$, then $p_{\mat{A}}
(\mat{A}) = \mat{0}$, where the polynomial is now evaluated with
$\mat{A}$ substituted for $\lambda$ and the constant term multiplied
by $\mat{I}$. For the $2 \times 2$ matrix of section 10.1 with
$p_{\mat{A}}(\lambda) = \lambda^{2} - 7\lambda + 10$, the theorem
asserts $\mat{A}^{2} - 7\mat{A} + 10\mat{I} = \mat{0}$, which the
reader can verify by direct multiplication. The diagonalisable
proof is immediate: $p_{\mat{A}}(\mat{A}) = \mat{S}\, p_{\mat{A}}
(\mat{\Lambda})\, \mat{S}\inv$, and $p_{\mat{A}}(\mat{\Lambda})$ is
diagonal with entries $p_{\mat{A}}(\lambda_{i}) = 0$, so the whole
product vanishes. The theorem holds for every square matrix,
including defective ones, by a continuity or adjugate argument.

The engineering payoff is that every power $\mat{A}^{k}$ for $k \geq
n$, and every matrix function expressible as a power series, reduces
to a polynomial in $\mat{A}$ of degree at most $n - 1$. The inverse,
when it exists, is one such polynomial: rearranging $\mat{A}^{n} +
c_{n-1}\mat{A}^{n-1} + \cdots + c_{1}\mat{A} + c_{0}\mat{I} =
\mat{0}$ and multiplying by $\mat{A}\inv$ gives $\mat{A}\inv =
-(1/c_{0})(\mat{A}^{n-1} + \cdots + c_{1}\mat{I})$, valid whenever
$c_{0} = (-1)^{n}\det\mat{A} \neq 0$. The relation is the basis of
the Krylov-subspace methods of Volume II Chapter 16: a matrix's
action is captured by its first $n$ powers acting on a vector.

\section{Quadratic forms and definiteness}\pathtag{core}

A symmetric matrix carries a function attached to it, the
\engterm{quadratic form} $q(\vect{x}) = \vect{x}\transpose \mat{A}
\vect{x}$. Quadratic forms are the energy of a vibrating system,
the curvature of an optimisation objective near a critical point,
the variance of a linear combination of random variables, and the
shape of a level set. The eigenvalues of $\mat{A}$ decide the sign
and the geometry of the form, and the vocabulary of definiteness
that follows is used in every optimisation argument in the rest of
the volume.

\subsection{Definiteness and the eigenvalue test}

A symmetric matrix $\mat{A}$ is

\begin{itemize}[leftmargin=*]
  \item \engterm{positive definite} if $q(\vect{x}) > 0$ for every
    nonzero $\vect{x}$;
  \item \engterm{positive semidefinite} if $q(\vect{x}) \geq 0$ for
    every $\vect{x}$;
  \item \engterm{negative definite} or \engterm{negative
    semidefinite} with the inequalities reversed;
  \item \engterm{indefinite} if $q$ takes both signs.
\end{itemize}

The eigenvalues settle the classification completely. In the
orthonormal eigenbasis, write $\vect{x} = \sum_{i} y_{i}
\vect{v}_{i}$; then $q(\vect{x}) = \sum_{i} \lambda_{i} y_{i}^{2}$,
a weighted sum of squares with the eigenvalues as weights. The form
is positive definite exactly when every $\lambda_{i} > 0$, positive
semidefinite when every $\lambda_{i} \geq 0$, negative definite when
every $\lambda_{i} < 0$, and indefinite when the eigenvalues
straddle zero. The geometry of the level set $q(\vect{x}) = 1$
follows at once: an ellipsoid with semi-axis lengths
$1/\sqrt{\lambda_{i}}$ along the eigenvectors when the form is
positive definite, a degenerate cylinder when one eigenvalue is
zero, a hyperboloid when the signs are mixed.

\subsection{Cheaper tests: pivots and leading minors}

Computing eigenvalues to test definiteness is more work than the
test needs. Two cheaper tests exist for a symmetric $\mat{A}$.
Sylvester's criterion: $\mat{A}$ is positive definite if and only
if all $n$ leading principal minors (the determinants of the
top-left $1 \times 1$, $2 \times 2$, \ldots, $n \times n$ blocks)
are positive. The pivot test: $\mat{A}$ is positive definite if and
only if Gaussian elimination without row exchanges produces $n$
positive pivots, which is exactly the condition that the Cholesky
factorisation $\mat{A} = \mat{L} \mat{L}\transpose$ succeeds.
Cholesky is the production test: it costs half the work of a general
$\mat{LU}$ factorisation and signals indefiniteness by encountering
a non-positive pivot mid-factorisation, at which point it stops.

\begin{archetype}[Scaling, in the form of curvature]
A quadratic form is a scaling of squared coordinates by the
eigenvalues, $q(\vect{x}) = \sum_{i} \lambda_{i} y_{i}^{2}$ in the
eigenbasis. The eigenvalues are the curvatures of the form along its
principal directions. A positive-definite form is a bowl that curves
upward in every direction; an indefinite form is a saddle. The
optimisation chapters read the eigenvalues of the Hessian to
classify a critical point: all positive is a local minimum, all
negative a local maximum, mixed signs a saddle. The same scaling
archetype that named eigenvalues in section 10.1 names the curvature
of the objective here.
\end{archetype}

\subsection{Worked example: classifying a quadratic form}

Consider $q(\vect{x}) = 5x_{1}^{2} + 8x_{1}x_{2} + 5x_{2}^{2}$. The
symmetric matrix is

\[
\mat{A} = \begin{pmatrix} 5 & 4 \\ 4 & 5 \end{pmatrix},
\]

with the off-diagonal $4$ being half the cross-term coefficient $8$,
because the cross term $2 A_{12} x_{1} x_{2}$ splits between the
$(1,2)$ and $(2,1)$ entries. The trace is $10$, the determinant is
$25 - 16 = 9$, and the eigenvalues solve $\lambda^{2} - 10\lambda +
9 = (\lambda - 9)(\lambda - 1)$, giving $\lambda_{1} = 9$ and
$\lambda_{2} = 1$. Both positive, so the form is positive definite.
The eigenvectors are $\vect{v}_{1} = (1, 1)/\sqrt{2}$ and
$\vect{v}_{2} = (1, -1)/\sqrt{2}$, so in the rotated coordinates
$y_{1} = (x_{1} + x_{2})/\sqrt{2}$, $y_{2} = (x_{1} - x_{2})/\sqrt{2}$
the form becomes $9 y_{1}^{2} + y_{2}^{2}$. The level set
$q = 1$ is an ellipse with semi-axes $1/3$ along $\vect{v}_{1}$ and
$1$ along $\vect{v}_{2}$: nine times stiffer along the diagonal than
along the anti-diagonal.

Sylvester's criterion confirms the classification without
eigenvalues: the leading minors are $5 > 0$ and $\det \mat{A} = 9 >
0$, both positive, so the form is positive definite. Change the
$(1,1)$ entry to $3$ and the determinant becomes $9 - 16 = -7 < 0$;
the second minor is negative, the form is indefinite, the level set
is a hyperbola, and the matrix has one positive and one negative
eigenvalue. The sign of a single $2 \times 2$ determinant flips the
geometry from bowl to saddle.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-quadratic-conics.tex}

\Cref{fig:vol02:ch10:quadratic-conics} draws the three geometries the
eigenvalue signs produce. Two positive eigenvalues give the ellipse of
the worked example, tilted along its principal directions with the
short axis along the stiff eigenvector; one sign flip gives the
hyperbola, its asymptotes running along the directions where the form
vanishes; a single zero eigenvalue collapses the ellipse into a pair
of parallel lines, flat along the null direction. The reader who
classifies a matrix by its eigenvalue signs is naming which of these
three shapes its energy surface takes.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-definiteness-map.tex}

\Cref{fig:vol02:ch10:definiteness-map} lifts Sylvester's criterion
into a single diagram in the trace-determinant plane. The determinant
sign draws the line between definite and indefinite matrices; among the
definite ones, the trace sign separates positive from negative. The
worked example's slide from bowl to saddle, driven by dropping the
$(1,1)$ entry from $5$ to $3$, is one horizontal step downward across
the $\delta = 0$ axis of that map: the determinant crosses zero and the
matrix changes character while the trace barely moves.

\begin{example}
The definiteness test is where a Hessian earns its keep. Consider
minimising $f(x_1, x_2) = x_1^{2} + x_1 x_2 + x_2^{2} - 3x_1$. The
gradient is $(\,2x_1 + x_2 - 3,\ x_1 + 2x_2\,)$, which vanishes at the
single critical point $x_1 = 2$, $x_2 = -1$. The Hessian is the
constant symmetric matrix
\[
\mat{H} = \begin{pmatrix} 2 & 1 \\ 1 & 2 \end{pmatrix},
\]
with trace $4$ and determinant $4 - 1 = 3$. Both leading minors are
positive ($2 > 0$ and $3 > 0$), so Sylvester's criterion certifies
$\mat{H}$ positive definite without computing an eigenvalue, and the
critical point is a strict local minimum. The eigenvalues, if wanted,
are $3$ and $1$, both positive, confirming the classification and
saying the objective curves three times as sharply along $(1,1)/
\sqrt{2}$ as along $(1,-1)/\sqrt{2}$: the same anisotropy that sets the
conditioning of a gradient-descent step on this bowl. Change the
cross-term coefficient from $1$ to $3$ and the Hessian becomes
$\begin{psmallmatrix} 2 & 3 \\ 3 & 2 \end{psmallmatrix}$ with
determinant $4 - 9 = -5 < 0$; the critical point turns into a saddle,
and a minimiser that trusted the vanishing gradient without checking
the Hessian sign would have stopped at a point that is not a minimum
at all.
\end{example}

\begin{estimation}
\textbf{Question.} A symmetric $10 \times 10$ stiffness matrix has
diagonal entries near $4$ and off-diagonal entries near $-1$, the
pattern of a one-dimensional chain. Estimate whether it is positive
definite, and estimate its smallest eigenvalue, without computing
the spectrum.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
Gerschgorin places every eigenvalue in a disc centred at $4$ with
radius at most $2$ (two neighbours, each contributing $\abs{-1} =
1$), so the spectrum lies in $[2, 6]$, comfortably positive. The
matrix is positive definite, and the smallest eigenvalue is at least
$2$. The exact tridiagonal chain has eigenvalues $4 - 2\cos(j\pi/11)$
for $j = 1, \ldots, 10$, the smallest being $4 - 2\cos(\pi/11)
\approx 4 - 2(0.96) \approx 2.08$, just inside the Gerschgorin bound.
The free Gerschgorin estimate of $2$ is within $5\%$ of the truth.
The diagonal dominance of the chain ($4 > 1 + 1$) was enough to
guarantee positive definiteness and a usable lower bound on the
stiffness, both without a single matrix operation.
\end{estimation}

\section{The singular value decomposition}\pathtag{core}

The diagonalisation of section 10.2 applies only to square matrices,
and only to those with a basis of eigenvectors. The singular value
decomposition (SVD) applies to every matrix, square or rectangular,
real or complex, full rank or rank deficient, and produces a
factorisation that is the engineer's first universal tool for
linear algebra.

\subsection{Existence and structure}

\begin{theorem}[Singular value decomposition]
Every matrix $\mat{A} \in \R^{m \times n}$ admits a factorisation

\[
\mat{A} = \mat{U} \mat{\Sigma} \mat{V}\transpose,
\]

where $\mat{U} \in \R^{m \times m}$ and $\mat{V} \in \R^{n \times n}$
are orthogonal matrices, and $\mat{\Sigma} \in \R^{m \times n}$ has
non-negative entries on its main diagonal and zeros elsewhere. The
diagonal entries $\sigma_{1} \geq \sigma_{2} \geq \cdots \geq
\sigma_{p} \geq 0$, where $p = \min(m, n)$, are the
\engterm{singular values} of $\mat{A}$. The number of nonzero
singular values equals $\rank \mat{A}$.
\end{theorem}

The proof is constructive and short. The Gram matrix $\mat{A}\transpose
\mat{A}$ is symmetric and positive semidefinite, so by the spectral
theorem it has an orthonormal eigenbasis $\{\vect{v}_{1}, \ldots,
\vect{v}_{n}\}$ with non-negative eigenvalues $\sigma_{i}^{2}$,
ordered $\sigma_{1}^{2} \geq \cdots \geq \sigma_{n}^{2} \geq 0$.
Define $\sigma_{i} = \sqrt{\sigma_{i}^{2}}$, and for each $i$ with
$\sigma_{i} > 0$ set $\vect{u}_{i} = \mat{A} \vect{v}_{i} / \sigma_{i}$.
The vectors $\{\vect{u}_{i}\}$ are orthonormal in $\R^{m}$. Extend
to an orthonormal basis $\{\vect{u}_{1}, \ldots, \vect{u}_{m}\}$ by
adding any orthonormal basis of the orthogonal complement of the
span. Stack the bases as columns of $\mat{U}$ and $\mat{V}$, and
verify $\mat{A} = \mat{U} \mat{\Sigma} \mat{V}\transpose$ entry by
entry. The full proof appears in any numerical-linear-algebra text
\cite[ch.~4]{text:trefethen-bau1997}.

\subsection{The four fundamental subspaces, revisited}

Chapter 9 introduced four subspaces attached to a matrix: the column
space and left null space in $\R^{m}$, the row space and null space
in $\R^{n}$. The SVD reads all four off the factorisation.

Let $r = \rank \mat{A}$. The first $r$ columns of $\mat{U}$ form an
orthonormal basis of $\col(\mat{A})$; the remaining $m - r$ columns
form an orthonormal basis of $\nul(\mat{A}\transpose)$ (the left
null space). The first $r$ columns of $\mat{V}$ form an orthonormal
basis of the row space $\col(\mat{A}\transpose)$; the remaining
$n - r$ columns form an orthonormal basis of $\nul(\mat{A})$. The
SVD, in one factorisation, gives orthonormal bases for all four
fundamental subspaces simultaneously, ordered by how strongly the
matrix uses or ignores each direction.

\subsection{Geometric picture}

The geometric reading of the SVD is that every linear map factors
into three operations: a rotation, an axis-aligned scaling, and a
second rotation. Specifically, $\mat{V}\transpose$ rotates the input
to align with the matrix's preferred directions; $\mat{\Sigma}$
scales each axis by the corresponding singular value; $\mat{U}$
rotates the result to align with the output basis. The unit ball in
$\R^{n}$ maps to an ellipsoid in $\R^{m}$ whose semi-axes are
$\sigma_{1} \vect{u}_{1}, \ldots, \sigma_{r} \vect{u}_{r}$. The
ellipsoid degenerates in $n - r$ directions (those mapped to zero)
and lives in the $r$-dimensional subspace $\col(\mat{A})$.

The picture is exact for every matrix. A symmetric positive-definite
$\mat{A}$ has $\mat{U} = \mat{V}$ and $\mat{\Sigma} = \mat{\Lambda}$;
the rotation before equals the rotation after; the matrix is its own
spectral decomposition. A general matrix needs two distinct
rotations, and the two are different exactly when the action of the
matrix is not symmetric. The reader should hold this picture in mind
through the chapter; every later derivation follows it.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-svd-geometry.tex}

\Cref{fig:vol02:ch10:svd-geometry} sets out the three stages
explicitly. The unit ball arrives, gets rotated to align with the
matrix's preferred input directions, gets stretched along each axis
by its singular value, and gets rotated again to align with the
output basis. The output is an ellipsoid whose semi-axis lengths are
exactly the singular values and whose semi-axis directions are
exactly the left singular vectors. For a rectangular matrix the
diagram is the same with the ambient dimension changing; for a
rank-deficient matrix the ellipsoid degenerates in some directions
(those mapped to zero); for a near-singular matrix one semi-axis is
short and the condition number is its aspect ratio.

\subsection{Computation and conditioning}

The cost of computing the SVD of an $m \times n$ matrix with $m \geq
n$ is $\oom{mn^{2} + n^{3}}$ flops. For a square matrix this is
$\oom{n^{3}}$, comparable to LU and twice QR. The standard algorithm
(Golub-Reinsch, 1970) reduces $\mat{A}$ to bidiagonal form by
Householder reflections and then iterates a variant of the QR
algorithm on the bidiagonal until the off-diagonal entries vanish
\cite[ch.~31]{text:trefethen-bau1997}. The algorithm is backward
stable. The computed singular values are exact for a nearby matrix
$\mat{A} + \mat{E}$ with $\norm{\mat{E}}_{2}/\norm{\mat{A}}_{2}$ on
the order of machine precision.

The condition number of a matrix in the 2-norm, introduced in
Chapter 9, is the ratio $\kappa(\mat{A}) = \sigma_{1}/\sigma_{r}$
of largest to smallest nonzero singular value. The SVD makes the
condition number visible. A matrix with $\sigma_{1} = 10$ and
$\sigma_{r} = 10^{-10}$ has $\kappa = 10^{11}$ and stretches some
inputs by $10$ while shrinking others to $10^{-10}$ of their length.
The conditioning trouble of Chapter 9 is the same trouble read
off the spectrum directly.

\subsection{Worked example: an SVD by hand}

The constructive proof becomes concrete on a small matrix. Consider

\[
\mat{A} = \begin{pmatrix} 1 & 1 \\ 0 & 1 \\ 1 & 0 \end{pmatrix}
\in \R^{3 \times 2}.
\]

The first step is the Gram matrix

\[
\mat{A}\transpose \mat{A} = \begin{pmatrix} 1 & 0 & 1 \\ 1 & 1 & 0
\end{pmatrix} \begin{pmatrix} 1 & 1 \\ 0 & 1 \\ 1 & 0 \end{pmatrix}
= \begin{pmatrix} 2 & 1 \\ 1 & 2 \end{pmatrix}.
\]

This $2 \times 2$ symmetric matrix has trace $4$, determinant $3$,
and eigenvalues solving $\lambda^{2} - 4\lambda + 3 = (\lambda - 3)
(\lambda - 1)$, so $\sigma_{1}^{2} = 3$ and $\sigma_{2}^{2} = 1$.
The singular values are $\sigma_{1} = \sqrt{3}$ and $\sigma_{2} = 1$.
The right singular vectors are the eigenvectors of $\mat{A}\transpose
\mat{A}$: at eigenvalue $3$, $\vect{v}_{1} = (1, 1)/\sqrt{2}$; at
eigenvalue $1$, $\vect{v}_{2} = (1, -1)/\sqrt{2}$.

The left singular vectors come from $\vect{u}_{i} = \mat{A}
\vect{v}_{i}/\sigma_{i}$:

\[
\vect{u}_{1} = \frac{1}{\sqrt{3}} \mat{A} \frac{(1,1)\transpose}
{\sqrt{2}} = \frac{1}{\sqrt{6}} \begin{pmatrix} 2 \\ 1 \\ 1
\end{pmatrix}, \qquad
\vect{u}_{2} = \frac{1}{1} \mat{A} \frac{(1,-1)\transpose}{\sqrt{2}}
= \frac{1}{\sqrt{2}} \begin{pmatrix} 0 \\ -1 \\ 1 \end{pmatrix}.
\]

Both are unit vectors and $\vect{u}_{1}\transpose \vect{u}_{2} = 0$,
as the construction guarantees. The matrix is $3 \times 2$ of rank
two, so the column space is two-dimensional and the third left
singular vector spans the one-dimensional left null space; it is the
unit normal to the plane spanned by the columns,
$\vect{u}_{3} = (1, -1, -1)/\sqrt{3}$ up to sign. The full
factorisation is $\mat{A} = \mat{U} \mat{\Sigma} \mat{V}\transpose$
with

\[
\mat{U} = \begin{pmatrix} 2/\sqrt{6} & 0 & 1/\sqrt{3} \\
1/\sqrt{6} & -1/\sqrt{2} & -1/\sqrt{3} \\
1/\sqrt{6} & 1/\sqrt{2} & -1/\sqrt{3} \end{pmatrix}, \quad
\mat{\Sigma} = \begin{pmatrix} \sqrt{3} & 0 \\ 0 & 1 \\ 0 & 0
\end{pmatrix}, \quad
\mat{V} = \frac{1}{\sqrt{2}} \begin{pmatrix} 1 & 1 \\ 1 & -1
\end{pmatrix}.
\]

The condition number is $\sigma_{1}/\sigma_{2} = \sqrt{3} \approx
1.73$, so the matrix is well conditioned. The reader who multiplies
$\mat{U} \mat{\Sigma} \mat{V}\transpose$ back out recovers $\mat{A}$
exactly. This hand computation is the algorithm production code
avoids (forming $\mat{A}\transpose \mat{A}$ squares the
conditioning, section 10.3's exercises), but it makes the geometry
of the factorisation visible: a $3 \times 2$ map sends the unit
circle of $\R^{2}$ to an ellipse living in a tilted plane of
$\R^{3}$, with semi-axes $\sqrt{3}\vect{u}_{1}$ and $\vect{u}_{2}$.

\subsection{The pseudoinverse and least squares}

The SVD gives the cleanest construction of the
\engterm{Moore-Penrose pseudoinverse}, the matrix that solves the
least-squares problem for any matrix, square or rectangular, full
rank or not. For $\mat{A} = \mat{U} \mat{\Sigma} \mat{V}\transpose$,
the pseudoinverse is

\[
\mat{A}^{+} = \mat{V} \mat{\Sigma}^{+} \mat{U}\transpose,
\]

where $\mat{\Sigma}^{+}$ is formed from $\mat{\Sigma}$ by
transposing it and replacing each nonzero singular value
$\sigma_{i}$ by its reciprocal $1/\sigma_{i}$, leaving the zeros in
place. The vector $\vect{x} = \mat{A}^{+} \vect{b}$ is the
minimum-norm least-squares solution of $\mat{A} \vect{x} =
\vect{b}$: among all $\vect{x}$ that minimise $\norm{\mat{A}\vect{x}
- \vect{b}}_{2}$, it is the one of smallest norm. When $\mat{A}$ is
square and invertible, $\mat{A}^{+} = \mat{A}\inv$; when $\mat{A}$
is tall with full column rank, $\mat{A}^{+} = (\mat{A}\transpose
\mat{A})\inv \mat{A}\transpose$, the normal-equations solution of
Chapter 9.

The pseudoinverse exposes the conditioning trap of least squares
directly. Each singular direction is scaled by $1/\sigma_{i}$, so a
tiny singular value $\sigma_{i}$ amplifies the component of
$\vect{b}$ along $\vect{u}_{i}$ by the large factor $1/\sigma_{i}$.
Measurement noise in $\vect{b}$, projected onto the small-singular-
value directions, is magnified into large, meaningless swings in
$\vect{x}$. The remedy is the \engterm{truncated pseudoinverse}:
discard the singular directions below a threshold, treating their
reciprocals as zero rather than as enormous. This is regularisation
read off the spectrum, and it is the bridge to Tikhonov
regularisation and ridge regression in Volume VII.

\begin{example}
The minimum-norm property is visible on a rank-deficient system where
ordinary least squares has no unique answer. Take
\[
\mat{A} = \begin{pmatrix} 1 & 1 \\ 1 & 1 \end{pmatrix}, \qquad
\vect{b} = \begin{pmatrix} 2 \\ 4 \end{pmatrix}.
\]
The matrix has rank one, so the columns span only the line through
$(1,1)$, and $\mat{A}\vect{x} = \vect{b}$ has no exact solution because
$\vect{b}$ is off that line. Least squares projects $\vect{b}$ onto the
column space: the projection is the average, $(3, 3)\transpose$, so the
best achievable output is $3$ in both rows and the residual is
$\norm{\vect{b} - (3,3)\transpose}_2 = \norm{(-1,1)\transpose}_2 =
\sqrt{2}$. Any $\vect{x} = (x_1, x_2)$ with $x_1 + x_2 = 3$ achieves
this residual, an entire line of least-squares solutions. The
pseudoinverse picks the one of smallest norm. The SVD is $\sigma_1 =
2$ with $\vect{u}_1 = \vect{v}_1 = (1,1)/\sqrt{2}$ and $\sigma_2 = 0$,
so
\[
\mat{A}^{+} = \frac{1}{\sigma_1}\vect{v}_1\vect{u}_1\transpose
= \frac{1}{2}\cdot\frac{1}{2}\begin{pmatrix} 1 & 1 \\ 1 & 1
\end{pmatrix} = \begin{pmatrix} 1/4 & 1/4 \\ 1/4 & 1/4 \end{pmatrix},
\]
and $\vect{x} = \mat{A}^{+}\vect{b} = (1.5, 1.5)\transpose$. This is
the point on the line $x_1 + x_2 = 3$ closest to the origin, the
minimum-norm least-squares solution, and the zero singular value is
handled by leaving its reciprocal at zero rather than blowing up. Had
the engineer instead formed the normal equations $\mat{A}\transpose
\mat{A}\vect{x} = \mat{A}\transpose\vect{b}$, the matrix
$\mat{A}\transpose\mat{A} = \begin{psmallmatrix} 2 & 2 \\ 2 & 2
\end{psmallmatrix}$ is singular and the solve fails outright; the
pseudoinverse route returns the sensible answer where the normal
equations refuse to run.
\end{example}

\subsection{The polar decomposition}\pathtag{mastery}

A square matrix factors as $\mat{A} = \mat{Q} \mat{P}$, with
$\mat{Q}$ orthogonal and $\mat{P}$ symmetric positive semidefinite,
the \engterm{polar decomposition}. It is the matrix analogue of
writing a complex number $z = e^{i\theta} r$ as a rotation times a
non-negative stretch. The SVD delivers it immediately: from
$\mat{A} = \mat{U} \mat{\Sigma} \mat{V}\transpose$, group as
$\mat{A} = (\mat{U} \mat{V}\transpose)(\mat{V} \mat{\Sigma}
\mat{V}\transpose) = \mat{Q} \mat{P}$, where $\mat{Q} = \mat{U}
\mat{V}\transpose$ is orthogonal and $\mat{P} = \mat{V} \mat{\Sigma}
\mat{V}\transpose$ is symmetric positive semidefinite. The
decomposition is the standard apparatus of continuum mechanics: the
deformation gradient of a strained body factors into a rotation
$\mat{Q}$ and a stretch tensor $\mat{P}$, separating rigid rotation
from genuine deformation. It also solves the orthogonal Procrustes
problem (find the rotation that best aligns two point sets), which
recurs in computer vision, crystallography, and molecular
structure alignment.

\section{Spectral methods and low-rank approximation}\pathtag{core}

The SVD orders the singular values from largest to smallest. Often
the largest few carry most of the action and the rest are noise or
redundancy. Truncating to the largest $k$ produces the best rank-$k$
approximation to the matrix, in a sense the Eckart-Young theorem
makes precise. This is the formal home of dimensionality reduction.

\subsection{The Eckart-Young theorem}

\begin{theorem}[Eckart-Young]
Let $\mat{A} = \mat{U} \mat{\Sigma} \mat{V}\transpose$ be the SVD of
$\mat{A} \in \R^{m \times n}$. For any integer $k$ with $1 \leq k \leq
\rank \mat{A}$, the matrix

\[
\mat{A}_{k} = \sum_{i=1}^{k} \sigma_{i} \vect{u}_{i} \vect{v}_{i}\transpose
\]

is the rank-$k$ matrix closest to $\mat{A}$ in both the spectral
norm and the Frobenius norm:

\[
\min_{\rank \mat{B} \leq k} \norm{\mat{A} - \mat{B}}_{2} =
\norm{\mat{A} - \mat{A}_{k}}_{2} = \sigma_{k+1},
\]

\[
\min_{\rank \mat{B} \leq k} \norm{\mat{A} - \mat{B}}_{F}^{2} =
\norm{\mat{A} - \mat{A}_{k}}_{F}^{2} = \sum_{i=k+1}^{p} \sigma_{i}^{2}.
\]
\end{theorem}

The theorem says that the truncated SVD is optimal as a low-rank
approximation under either of the two engineering-relevant matrix
norms. The proof is a Courant-Fischer argument
\cite[ch.~5]{text:trefethen-bau1997}. The approximation error is
the next singular value (in spectral norm) or the tail energy (in
Frobenius norm), exactly. There is no constant or asymptotic in the
statement; it is sharp.

The truncated SVD is the engineering reading of \emph{the matrix is
mostly the first $k$ singular components}. When the singular values
decay quickly, $\mat{A}_{k}$ is close to $\mat{A}$ at small $k$ and
the matrix is effectively low-rank. When they decay slowly,
truncation costs accuracy and full rank is the genuine signal.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-eckart-young-error.tex}

\Cref{fig:vol02:ch10:eckart-young-error} plots the two error norms
against the retained rank for a matrix with a geometric spectrum. The
theorem's sharpness is what makes the plot a design tool rather than a
sketch: the spectral-norm curve is the next singular value exactly, and
the Frobenius-norm curve is the square root of the discarded tail
exactly, so the engineer reads the smallest rank that clears a
tolerance off the graph with no safety factor for slop in the bound
itself. The two norms answer different questions, and they usually
disagree by a component, the spectral norm asking only about the
largest thing thrown away and the Frobenius norm about the total energy
thrown away.

\begin{example}
The two error norms diverge most where they matter. Take a matrix with
singular values $\sigma = (10,\ 9,\ 8,\ 1,\ 1,\ 1,\ 1,\ 1,\ 1,\ 1)$, a
clean three-component signal on a flat tail of seven small values. At
rank $k = 3$ the spectral-norm error is $\sigma_{4} = 1$, a single
small value, and the plot's blue curve says the truncation looks nearly
perfect. The Frobenius-norm error tells a different story: it is
$\sqrt{\sigma_{4}^{2} + \cdots + \sigma_{10}^{2}} = \sqrt{7} \approx
2.65$, larger than the largest discarded value because the seven small
tail terms accumulate. Reported as a fraction of the total energy
$\sqrt{\sum_i \sigma_i^2} = \sqrt{100 + 81 + 64 + 7} = \sqrt{252}
\approx 15.9$, the rank-three Frobenius error is $2.65/15.9 \approx
17\%$, not the near-zero the spectral norm suggests. An engineer who
certifies the truncation on spectral-norm error alone would overstate
its quality whenever the tail is long, which is the natural-image case;
the honest report quotes both norms, and the scree plot below is how
the reader inspects the tail that the Frobenius norm sums.
\end{example}

\subsection{PCA from the SVD}

\engterm{Principal Component Analysis} is the SVD applied to a
mean-centred data matrix. Let $\mat{X} \in \R^{m \times n}$ have
$m$ samples in rows and $n$ features in columns. Centre by
subtracting the column mean: $\mat{X}_{c} = \mat{X} - \vect{1}
\bar{\vect{x}}\transpose$, where $\bar{\vect{x}}$ is the column-wise
mean. The sample covariance matrix is $\mat{C} = \mat{X}_{c}\transpose
\mat{X}_{c}/(m-1)$, an $n \times n$ symmetric positive-semidefinite
matrix.

The eigenvectors of $\mat{C}$ are the \engterm{principal directions}.
The eigenvalues are the variances captured along each principal
direction. The standard recipe computes the SVD of $\mat{X}_{c} =
\mat{U} \mat{\Sigma} \mat{V}\transpose$ and reads:

\begin{itemize}[leftmargin=*]
  \item The columns of $\mat{V}$ are the principal directions in the
    feature space.
  \item The squared singular values $\sigma_{i}^{2}/(m-1)$ are the
    variances along the principal directions.
  \item The matrix $\mat{X}_{c} \mat{V} = \mat{U} \mat{\Sigma}$
    gives the principal-component scores: each row is a sample
    expressed in the principal-direction basis.
\end{itemize}

The reduction to $k$ components keeps the first $k$ columns of
$\mat{V}$ and produces $\mat{X}_{c} \mat{V}_{k}$, an $m \times k$
matrix that is the best $k$-dimensional linear summary of the data
under the Eckart-Young criterion. The fraction of variance retained
is $\sum_{i=1}^{k} \sigma_{i}^{2} / \sum_{i=1}^{p} \sigma_{i}^{2}$,
the \engterm{explained-variance ratio}.

The numerical caution from Chapter 9 carries over. Forming $\mat{C} =
\mat{X}_{c}\transpose \mat{X}_{c}$ explicitly squares the conditioning
of $\mat{X}_{c}$. The SVD of $\mat{X}_{c}$ produces the same
principal directions without the squaring and is the reportable
algorithm. Section 10.8's worked PCA follows this discipline.

\subsection{Other spectral applications}

The truncated SVD recurs throughout engineering practice. In image
compression (section 10.8) the matrix is the pixel array and a
moderate $k$ recovers a recognisable image at a fraction of the
storage. In recommender systems the matrix is the
user-by-item ratings and the truncated SVD recovers latent factors.
In graph signal processing the matrix is a graph Laplacian and the
truncated eigenvectors form a Fourier basis adapted to the graph
structure. The common pattern is that data, when arranged as a
matrix, often have a few directions that carry most of the
variation. The SVD finds those directions and the truncation
projects onto them. The data-mining text by \textcite{text:v2ch10-elden2007}
covers the same pattern across information retrieval, image
processing, and microarray analysis with the bookkeeping in this
chapter's vocabulary.

The graph-Laplacian case deserves a closer look because it turns a
combinatorial problem into a spectral one. For a graph with adjacency
matrix $\mat{W}$ and diagonal degree matrix $\mat{D}$ (each diagonal
entry the sum of a row of $\mat{W}$), the \engterm{graph Laplacian} is
$\mat{L} = \mat{D} - \mat{W}$, symmetric and positive semidefinite. Its
smallest eigenvalue is always zero, with the all-ones eigenvector,
because every row of $\mat{L}$ sums to zero by construction. The
information lives in the next eigenvalues. The multiplicity of the zero
eigenvalue counts the connected components of the graph exactly: a
graph in two disconnected pieces has a two-dimensional null space
spanned by the indicator vectors of the two pieces. When the graph is
connected but nearly splits into two clusters joined by a few edges,
the second-smallest eigenvalue is small but nonzero, and its
eigenvector, the \engterm{Fiedler vector}, is positive on one cluster
and negative on the other. Thresholding the Fiedler vector at zero
partitions the graph into its two natural clusters, an approximate
minimum-cut computed by one eigenvector rather than by searching over
exponentially many partitions. This is \engterm{spectral clustering},
and its correctness rests on the same spectral-gap reasoning as power
iteration: a clear gap between the second and third Laplacian
eigenvalues means the two-cluster structure is real, and no gap means
the graph has no clean two-way split to find. The eigenvalues of a
matrix built from a graph's connectivity read off the graph's
large-scale shape, which is why the same apparatus serves image
segmentation, community detection in networks, and the reordering that
makes a sparse matrix cheaper to factor.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-singular-value-decay.tex}

\Cref{fig:vol02:ch10:singular-value-decay} compares four schematic
decay curves. The engineer in front of a new matrix runs the SVD,
plots $\sigma_{k}/\sigma_{1}$ on a log-y axis against $k$, and reads
off where the curve crosses the application's acceptable error level.
A sharp drop at small $k$ followed by a plateau is the
low-rank-plus-noise pattern: the effective rank is well defined and
aggressive truncation is safe. A power-law decay (the natural-image
case, blue dashed) gives a gradual quality-storage trade with no
sharp threshold. A nearly flat curve (the random-Gaussian case,
amber) says no useful low-rank approximation exists and the matrix
has to be kept at full rank.

\begin{estimation}
\textbf{Question.} A grayscale photograph is $2000 \times 3000$
pixels. The reader wants a storage saving of at least $10\times$
through truncated-SVD compression. Estimate the rank $k$ at which
this storage saving is achieved, and judge whether the resulting
reconstruction is likely to be acceptable for the natural-image
class.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The original storage is $mn = 2000 \times 3000 = 6 \times 10^{6}$
entries. The truncated-SVD storage is $k(m + 1 + n) \approx 5000 k$
entries, since $m + n = 5000$. A tenfold saving demands
$5000 k \leq 6 \times 10^{5}$, or $k \leq 120$. At $k = 120$, the
ratio $\sigma_{120}/\sigma_{1}$ for a natural image with $\sigma_{k}
\propto 1/k$ is roughly $1/120 \approx 0.008$, below the $1\%$
threshold in
\cref{fig:vol02:ch10:singular-value-decay}. The reconstruction is
likely acceptable for casual viewing but will lose fine texture and
edges. Section 10.8's image-compression worked example confirms the
threshold experimentally on a specific image. The same back-of-the-
envelope works for any compression ratio $r$: solve $k(m+n) \leq
mn/r$ for $k$ and check the spectrum at that $k$.
\end{estimation}

\subsection{Latent semantic analysis (recognition)}

Latent semantic analysis (LSA) applies the truncated SVD to a
term-by-document matrix to recover topical structure. The columns
of $\mat{V}_{k}$ form a basis of \emph{topics}; the rows of
$\mat{U}_{k}$ map terms onto topics; the rows of $\mat{X}_{c}
\mat{V}_{k}$ map documents onto topics. The technique is from the
late 1980s and is foundational for the embedding methods of Volume
VII (information). The engineering payoff is that two documents sharing
no words can still land close in the $k$-dimensional topic space if
they use synonymous vocabulary that co-occurs with the same third
terms: the truncation folds synonyms together because they load on the
same singular direction, and it separates the two senses of a
polysemous word because they load on different ones. A retrieval query,
projected into the same topic space by $\vect{q}\mat{V}_k$, then matches
documents by topic overlap rather than by literal word overlap, which
is why LSA retrieved relevant documents that keyword search missed.
The choice of $k$ is the same scree-plot judgment as everywhere else in
this section: too small and distinct topics collapse together, too
large and the retained tail is sampling noise in the word counts. We
mention the method at recognition level. The reader who completes the
project of this chapter has implemented PCA, which is the same
truncated-SVD computation on a mean-centred numerical data matrix
rather than a term-by-document count matrix.

\subsection{Randomised SVD for large low-rank matrices}\pathtag{mastery}

The Golub-Reinsch SVD costs $\oom{mn^{2}}$, which is prohibitive when
the engineer needs only the top $k$ singular components of a matrix
with $m$ and $n$ in the millions. When the matrix is well
approximated by a rank-$k$ object, a randomised algorithm computes the
truncated SVD in time proportional to $k$ rather than to the smaller
dimension, and the same Eckart-Young theorem certifies the result.

The idea is to find a small basis that captures the matrix's range.
Draw a random Gaussian matrix $\mat{\Omega} \in \R^{n \times (k + p)}$
with a small oversampling $p$ (typically $p = 5$ to $10$), and form
$\mat{Y} = \mat{A}\mat{\Omega}$. Because $\mat{\Omega}$ mixes the
columns of $\mat{A}$ at random, the columns of $\mat{Y}$ almost surely
span the dominant left-singular subspace of $\mat{A}$. Orthonormalise
$\mat{Y}$ by a QR factorisation to get $\mat{Q}$ with $k + p$
orthonormal columns; then $\mat{A} \approx \mat{Q}\mat{Q}\transpose
\mat{A}$ to high accuracy. Form the small matrix $\mat{B} =
\mat{Q}\transpose \mat{A} \in \R^{(k+p) \times n}$, take its SVD
$\mat{B} = \tilde{\mat{U}}\mat{\Sigma}\mat{V}\transpose$ (cheap,
because $\mat{B}$ is short), and set $\mat{U} = \mat{Q}\tilde{\mat{U}}$.
The result is a rank-$(k + p)$ SVD of $\mat{A}$ whose top $k$
components are accurate to a factor close to the Eckart-Young optimum.

\begin{principle}[Randomised range finding]
For a matrix $\mat{A}$ whose singular values decay, the randomised SVD
with oversampling $p$ produces a rank-$k$ approximation whose expected
spectral-norm error is bounded by a small multiple of the optimal
$\sigma_{k+1}$, with the multiple controlled by $p$ and shrinking
fast as $p$ grows. One or two \engterm{power-iteration} passes,
replacing $\mat{Y} = \mat{A}\mat{\Omega}$ by $\mat{Y} =
(\mat{A}\mat{A}\transpose)^{q}\mat{A}\mat{\Omega}$, sharpen the
accuracy when the singular values decay slowly, at the cost of $q$
extra matrix-products.
\end{principle}

The cost is $\oom{mn\log k}$ for the random projection plus
$\oom{(m + n)k^{2}}$ for the small dense work, against $\oom{mn^{2}}$
for the full SVD. For a matrix that is genuinely close to rank $k$,
the saving is enormous, and the method is the workhorse behind
large-scale PCA, recommender-system factorisation, and
reduced-order modelling \cite[ch.~4]{text:v2ch10-golub-vanloan2013}.
The honesty check is the same scree plot as always: a randomised SVD
applied to a matrix with no low-rank structure (the slow tail of a
natural image, the flat spectrum of Gaussian noise) returns a poor
approximation, and the error bound degrades to match. Randomisation
buys speed only where the structure the truncated SVD assumes is
actually present.

\subsection{QR decomposition and the QR algorithm}\pathtag{standard}

Every matrix $\mat{A} \in \R^{m \times n}$ with $m \geq n$ factors
as $\mat{A} = \mat{Q} \mat{R}$, with $\mat{Q} \in \R^{m \times n}$
having orthonormal columns and $\mat{R} \in \R^{n \times n}$ upper
triangular. The columns of $\mat{Q}$ are an orthonormal basis for
the column space of $\mat{A}$, built by Gram-Schmidt
orthogonalisation of the columns of $\mat{A}$, and $\mat{R}$ records
the coefficients that express each original column in that basis.
Production code uses Householder reflections rather than classical
Gram-Schmidt, because the classical procedure loses orthogonality
under rounding while the Householder version is backward stable
\cite[ch.~10]{text:trefethen-bau1997}.

The QR decomposition is the engineer's default for overdetermined
least squares: to solve $\min \norm{\mat{A}\vect{x} - \vect{b}}_{2}$
with $\mat{A}$ tall and full column rank, factor $\mat{A} = \mat{Q}
\mat{R}$ and solve $\mat{R} \vect{x} = \mat{Q}\transpose \vect{b}$
by back substitution. The route avoids forming $\mat{A}\transpose
\mat{A}$ and the conditioning squaring that comes with it, the same
discipline the SVD imposes, at lower cost than the SVD when the full
spectral picture is not needed.

\begin{example}
Gram-Schmidt makes the factorisation concrete. Take
\[
\mat{A} = \begin{pmatrix} 1 & 1 \\ 1 & 0 \\ 0 & 1 \end{pmatrix},
\]
with columns $\vect{a}_1 = (1,1,0)\transpose$ and $\vect{a}_2 =
(1,0,1)\transpose$. Normalise the first column: $\norm{\vect{a}_1} =
\sqrt{2}$, so $\vect{q}_1 = (1,1,0)\transpose/\sqrt{2}$ and $R_{11} =
\sqrt{2}$. Project the second column off $\vect{q}_1$: the overlap is
$R_{12} = \vect{q}_1\transpose\vect{a}_2 = (1 + 0 + 0)/\sqrt{2} =
1/\sqrt{2}$, and the residual is $\vect{a}_2 - R_{12}\vect{q}_1 =
(1,0,1) - \tfrac{1}{2}(1,1,0) = (1/2, -1/2, 1)$. Its norm is
$\sqrt{1/4 + 1/4 + 1} = \sqrt{3/2}$, so $R_{22} = \sqrt{3/2}$ and
$\vect{q}_2 = (1/2, -1/2, 1)\transpose/\sqrt{3/2}$. The factorisation is
\[
\mat{Q} = \begin{pmatrix} 1/\sqrt{2} & 1/\sqrt{6} \\ 1/\sqrt{2} &
-1/\sqrt{6} \\ 0 & 2/\sqrt{6} \end{pmatrix}, \qquad
\mat{R} = \begin{pmatrix} \sqrt{2} & 1/\sqrt{2} \\ 0 & \sqrt{3/2}
\end{pmatrix},
\]
where $\vect{q}_2$ has been rewritten by clearing the common factor
($1/\sqrt{3/2}$ times $(1/2,-1/2,1)$ equals $(1,-1,2)/\sqrt{6}$). The
columns of $\mat{Q}$ are orthonormal, $\mat{R}$ is upper triangular
with the pivot norms on its diagonal, and multiplying $\mat{Q}\mat{R}$
recovers $\mat{A}$. To solve a least-squares problem
$\min\norm{\mat{A}\vect{x} - \vect{b}}_2$ the engineer then forms
$\mat{Q}\transpose\vect{b}$ and back-substitutes through $\mat{R}$,
never touching the ill-conditioned $\mat{A}\transpose\mat{A} =
\begin{psmallmatrix} 2 & 1 \\ 1 & 2 \end{psmallmatrix}$ whose formation
would have squared the conditioning. Classical Gram-Schmidt as written
here loses orthogonality under rounding for nearly dependent columns,
which is why production code runs the algebraically equivalent
Householder version, but the arithmetic is the same factorisation.
\end{example}

The QR decomposition also names the standard general eigenvalue
algorithm. The \engterm{QR algorithm} computes eigenvalues by
repeated factor-and-reverse: factor $\mat{A}_{k} = \mat{Q}_{k}
\mat{R}_{k}$, form $\mat{A}_{k+1} = \mat{R}_{k} \mat{Q}_{k}$, and
iterate. Each step is a similarity transformation ($\mat{A}_{k+1} =
\mat{Q}_{k}\transpose \mat{A}_{k} \mat{Q}_{k}$), so the eigenvalues
are preserved, and under mild conditions $\mat{A}_{k}$ converges to
an upper-triangular (or quasi-triangular) matrix whose diagonal
displays the eigenvalues. With shifts and a preliminary reduction to
Hessenberg form the algorithm is the workhorse behind LAPACK's dense
eigensolvers and is what a library call to compute eigenvalues
actually runs \cite[ch.~28-29]{text:trefethen-bau1997}. The reader
should recognise that the eigenvalue computation the chapter has
been quoting as a black box is itself built from the QR
decomposition, applied iteratively, with the Schur form of section
10.7 as its limit.

\section{Generalised eigenproblems and matrix functions}\pathtag{standard}

The eigenproblem $\mat{A}\vect{x} = \lambda\vect{x}$ is the special
case of a wider family. Physical models routinely produce a second
matrix on the right-hand side, $\mat{A}\vect{x} = \lambda\mat{B}\vect{x}$,
and the engineer who forces such a problem into the standard form by
inverting $\mat{B}$ pays in conditioning and in lost symmetry. Matrix
functions, the exponential above all, sit in the same family: once the
spectral decomposition is in hand, any scalar function lifts to the
matrix. The generalised eigenproblem, simultaneous diagonalisation,
matrix functions, and the large-scale and random-matrix tools that an
engineering analysis reaches for once the basic eigenproblem is
understood all follow.

\subsection{The generalised eigenproblem}

A \engterm{generalised eigenvalue problem} is the equation
$\mat{A}\vect{x} = \lambda\mat{B}\vect{x}$ for a pair of $n \times n$
matrices $(\mat{A}, \mat{B})$. The scalars $\lambda$ for which a nonzero
$\vect{x}$ exists are the \engterm{generalised eigenvalues}, and they
are the roots of $\det(\mat{A} - \lambda\mat{B}) = 0$. When $\mat{B} =
\mat{I}$ the problem is the ordinary one. The pair $(\mat{A}, \mat{B})$
is called a \engterm{matrix pencil}, and its spectrum can include the
value $\infty$ (when $\mat{B}$ is singular and the leading coefficient
of the determinant polynomial drops) alongside the finite eigenvalues.

The setting that makes the generalised problem unavoidable is
structural vibration. The undamped equation of motion of a multi-degree
of freedom system is $\mat{M}\ddot{\vect{q}} + \mat{K}\vect{q} =
\vect{0}$, with $\mat{M}$ the mass matrix and $\mat{K}$ the stiffness
matrix, both symmetric and $\mat{M}$ positive definite. Seeking
harmonic solutions $\vect{q}(t) = \vect{x}\,e^{i\omega t}$ gives
$(\mat{K} - \omega^{2}\mat{M})\vect{x} = \vect{0}$, the generalised
eigenproblem $\mat{K}\vect{x} = \lambda\mat{M}\vect{x}$ with
$\lambda = \omega^{2}$. The natural frequencies are
$\omega_{i} = \sqrt{\lambda_{i}}$ and the mode shapes are the
generalised eigenvectors. The modal analysis of section 10.8 quietly
assumed $\mat{M} = m\mat{I}$, equal masses, which collapses the
generalised problem to an ordinary one. Unequal masses break that
collapse.

The temptation is to write $\mat{M}\inv\mat{K}\vect{x} = \lambda\vect{x}$
and call an ordinary eigensolver. The product $\mat{M}\inv\mat{K}$ is in
general not symmetric even though $\mat{M}$ and $\mat{K}$ both are, so
the orthogonality of mode shapes and the realness of the spectrum, both
guaranteed for the symmetric pencil, are thrown away and must be
recovered numerically. The disciplined route keeps the symmetry. Factor
$\mat{M} = \mat{L}\mat{L}\transpose$ by Cholesky (it succeeds because
$\mat{M}$ is positive definite, section 10.3), substitute
$\vect{x} = \mat{L}^{-\transpose}\vect{y}$, and the problem becomes
$\mat{L}\inv\mat{K}\mat{L}^{-\transpose}\vect{y} = \lambda\vect{y}$,
an ordinary eigenproblem whose matrix $\tilde{\mat{K}} =
\mat{L}\inv\mat{K}\mat{L}^{-\transpose}$ is symmetric. Its eigenvalues
are the generalised eigenvalues; its orthonormal eigenvectors
$\vect{y}_{i}$ map back to mode shapes $\vect{x}_{i} =
\mat{L}^{-\transpose}\vect{y}_{i}$ that are orthonormal in the
\engterm{$\mat{M}$-inner product} $\langle\vect{x}, \vect{x}'\rangle_{M}
= \vect{x}\transpose\mat{M}\vect{x}'$. Mode shapes are orthogonal with
respect to the mass, not with respect to the bare dot product, and the
$\mat{M}$-orthonormalisation is what lets a forced response decouple
into independent single-degree-of-freedom oscillators. Production code
calls the LAPACK symmetric-definite generalised driver \texttt{?SYGV},
which performs exactly this Cholesky reduction internally
\cite[ch.~8]{text:v2ch10-golub-vanloan2013}.

\subsection{Worked example: a two-mass system with unequal masses}

Two masses $m_{1} = 1$ and $m_{2} = 2$ sit in a line, joined to walls
and to each other by three springs of stiffness $k = 1$. The stiffness
and mass matrices are

\[
\mat{K} = \begin{pmatrix} 2 & -1 \\ -1 & 2 \end{pmatrix}, \qquad
\mat{M} = \begin{pmatrix} 1 & 0 \\ 0 & 2 \end{pmatrix}.
\]

The generalised eigenvalues solve $\det(\mat{K} - \lambda\mat{M}) = 0$:

\[
\det\begin{pmatrix} 2 - \lambda & -1 \\ -1 & 2 - 2\lambda \end{pmatrix}
= (2 - \lambda)(2 - 2\lambda) - 1 = 2\lambda^{2} - 6\lambda + 3 = 0,
\]

with roots $\lambda = (3 \pm \sqrt{3})/2$, that is $\lambda_{1} \approx
0.634$ and $\lambda_{2} \approx 2.366$. The natural frequencies are
$\omega_{1} = \sqrt{\lambda_{1}} \approx 0.796\sqrt{k/m_{1}}$ and
$\omega_{2} \approx 1.538\sqrt{k/m_{1}}$. The wrong shortcut, taking the
ordinary eigenvalues of $\mat{K}$ alone, would give $1$ and $3$ and
frequencies $1$ and $1.732$, both too high, because it pretends the
heavier mass weighs the same as the lighter one.

The mode shapes come from $(\mat{K} - \lambda_{i}\mat{M})\vect{x}_{i} =
\vect{0}$. At $\lambda_{1} \approx 0.634$ the first row reads
$(2 - 0.634)x_{1} - x_{2} = 0$, so $x_{2} = 1.366\,x_{1}$, a mode in
which both masses move the same way: in phase. At $\lambda_{2} \approx
2.366$ the first row reads $(2 - 2.366)x_{1} - x_{2} = 0$, so $x_{2} =
-0.366\,x_{1}$, a mode in which the masses move in opposition, with the
heavier $m_{2}$ carrying the smaller amplitude. The two mode shapes
satisfy $\vect{x}_{1}\transpose\mat{M}\vect{x}_{2} = 0$: taking
$\vect{x}_{1} = (1, 1.366)\transpose$ and $\vect{x}_{2} = (1,
-0.366)\transpose$, the $\mat{M}$-inner product is $1 \cdot 1 + 2 \cdot
1.366 \cdot (-0.366) = 1 - 1.0 \approx 0$, orthogonal in the mass
metric though not in the ordinary one ($\vect{x}_{1}\transpose
\vect{x}_{2} = 1 - 0.5 = 0.5 \neq 0$). \Cref{fig:vol02:ch10:generalized-modes}
draws the two shapes.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-generalized-modes.tex}

\subsection{Simultaneous diagonalisation}\pathtag{mastery}

The Cholesky reduction above does more than solve one problem. It
diagonalises both $\mat{K}$ and $\mat{M}$ at once. The change of basis
$\vect{q} = \mat{X}\vect{\eta}$, with $\mat{X}$ the matrix of
$\mat{M}$-orthonormal mode shapes, gives $\mat{X}\transpose\mat{M}
\mat{X} = \mat{I}$ and $\mat{X}\transpose\mat{K}\mat{X} =
\diag(\lambda_{1}, \ldots, \lambda_{n})$. The same congruence transform
sends the kinetic-energy form to the identity and the potential-energy
form to a diagonal of squared frequencies. The coordinates
$\vect{\eta}$ are the \engterm{modal coordinates}, and in them the
coupled system $\mat{M}\ddot{\vect{q}} + \mat{K}\vect{q} = \vect{0}$
becomes $n$ independent scalar oscillators $\ddot{\eta}_{i} +
\lambda_{i}\eta_{i} = 0$. This is the whole power of modal analysis:
a single congruence decouples an arbitrarily large linear vibration
problem into a list of one-dimensional ones.

The general theorem is that two symmetric matrices can be simultaneously
diagonalised by a congruence $\mat{X}\transpose(\cdot)\mat{X}$ whenever
one of them is positive definite. They can be simultaneously
diagonalised by an orthogonal similarity (the same $\mat{Q}$ on both
sides, $\mat{Q}\transpose(\cdot)\mat{Q}$) if and only if they commute,
$\mat{A}\mat{B} = \mat{B}\mat{A}$, which is the matrix statement that
two observables share an eigenbasis. The commuting case is the
linear-algebra core of why simultaneously measurable quantities in a
linear theory share principal directions, and it recurs whenever an
analysis hopes to read two coupled matrices in one common basis.

\subsection{Applied vignette: keeping a machine off resonance}

The generalised eigenproblem earns its place in a design decision that
recurs across mechanical engineering: a rotating machine must not run
near a natural frequency of its support, or the vibration amplifies
without bound. A pump and its motor, a combined mass, sit on an
elevated steel frame. The frame flexes in two coupled ways, a
side-to-sway and a rocking tilt, and the design question is whether the
pump's running speed, current as of 2025, falls dangerously close to
either natural frequency of the loaded frame.

Reduced to two degrees of freedom, the sway displacement $q_1$ and the
tilt angle scaled to a displacement $q_2$, the loaded frame has mass
and stiffness matrices
\[
\mat{M} = \begin{pmatrix} 400 & 0 \\ 0 & 250 \end{pmatrix}\ \text{kg},
\qquad
\mat{K} = \begin{pmatrix} 2.7 & -0.5 \\ -0.5 & 3.4 \end{pmatrix}
\times 10^{6}\ \text{N/m},
\]
with the off-diagonal stiffness the coupling between sway and tilt that
a single-degree-of-freedom hand estimate ignores. The natural
frequencies solve the generalised eigenproblem $\mat{K}\vect{x} =
\lambda\mat{M}\vect{x}$ with $\lambda = \omega^{2}$, and the
disciplined route of the previous subsections keeps the symmetry: the
frequencies are the eigenvalues of $\tilde{\mat{K}} = \mat{L}\inv\mat{K}
\mat{L}^{-\transpose}$ with $\mat{M} = \mat{L}\mat{L}\transpose$, not
of the non-symmetric $\mat{M}\inv\mat{K}$.

The generalised eigenvalues are $\lambda_1 = 6.40 \times 10^{3}$ and
$\lambda_2 = 1.39 \times 10^{4}\ \text{s}^{-2}$, so the natural
frequencies are $\omega_1 = \sqrt{\lambda_1} = 80.0\ \text{rad/s}$ and
$\omega_2 = 118\ \text{rad/s}$, that is $f_1 = 12.7\ \text{Hz}$ and
$f_2 = 18.8\ \text{Hz}$. The pump runs at $780$ revolutions per minute,
a forcing frequency of $780/60 = 13.0\ \text{Hz}$. That sits within
$3\%$ of the first natural frequency $f_1 = 12.7\ \text{Hz}$, squarely
inside the resonance danger band (the usual design rule keeps the
forcing frequency at least $20\%$ away from any natural frequency), so
the design as drawn will shake itself apart. The wrong shortcut, taking
the ordinary eigenvalues of $\mat{K}$ alone and dividing by a single
lumped mass, would have reported frequencies too high by the square
root of the mass ratio and might have declared the design safe.

The mode shapes say how to fix it. The generalised eigenvector at
$\lambda_1$ is $(1,\ 0.28)$ up to normalisation, dominated by the sway
coordinate, so the first mode is a sway-led motion; the second mode
$(0.17,\ 1)$ is tilt-led. Stiffening the frame against sway, adding a
diagonal brace that raises the $(1,1)$ stiffness entry, moves $f_1$
upward and away from the pump speed while barely touching the tilt-led
second mode. Recomputing the generalised eigenproblem with the $(1,1)$
stiffness raised to $4.0 \times 10^{6}\ \text{N/m}$ lifts $f_1$ to
$15.4\ \text{Hz}$, now $19\%$ above the forcing frequency and just
inside the design margin. The whole intervention, brace the sway
direction and recheck, is one recomputation of a two-by-two generalised
eigenproblem, and the mode shape is what told the engineer which of the
two stiffness entries to raise rather than stiffening blindly. The
$\mat{M}$-orthonormal mode shapes of the simultaneous-diagonalisation
result are also what let the forced response be checked mode by mode:
the pump's harmonic force projects onto each mode, and only the mode
whose frequency is near the forcing shows large amplification.

\subsection{Matrix functions through the spectrum}

Section 10.2 noted that $f(\mat{A}) = \mat{S}f(\mat{\Lambda})\mat{S}\inv$
for a diagonalisable $\mat{A}$ and any scalar $f$ defined on the
eigenvalues. The identity is worth carrying further, because it is how
an engineer computes the response of a linear system, the steady-state
of a diffusion, or the square root of a covariance, and because its
failure modes are exactly the conditioning traps of the chapter.

The square root of a symmetric positive-definite matrix is the workhorse
case. Given $\mat{C} = \mat{Q}\mat{\Lambda}\mat{Q}\transpose$ with
$\mat{\Lambda} = \diag(\lambda_{i})$ and every $\lambda_{i} > 0$, the
matrix $\mat{C}^{1/2} = \mat{Q}\diag(\sqrt{\lambda_{i}})\mat{Q}
\transpose$ is the unique symmetric positive-definite matrix whose
square is $\mat{C}$. It is the object that whitens correlated noise: if
a random vector has covariance $\mat{C}$, then $\mat{C}^{-1/2}$ applied
to it produces a vector with identity covariance, the standard
preconditioning step in estimation and in the generation of correlated
samples (draw white noise, multiply by $\mat{C}^{1/2}$). The Cholesky
factor $\mat{L}$ of $\mat{C}$ does the same job more cheaply but is not
symmetric; the symmetric square root is preferred where the symmetry
itself matters, as in interpolating between covariances or in the
geometry of the positive-definite cone.

\subsection{Applied vignette: whitening correlated sensor noise}

A concrete estimation problem shows the matrix square root doing work
that no scalar operation can. Three accelerometers mounted on one rigid
bracket measure what should be independent axes, but shared mounting
compliance and a common power-supply ripple correlate their noise. A
bench calibration, current as of 2025, with the bracket at rest,
records the noise covariance
\[
\mat{C} = \begin{pmatrix} 4 & 2 & 0 \\ 2 & 5 & 2 \\ 0 & 2 & 4
\end{pmatrix}
\]
in units of $(\text{milli-}g)^{2}$, symmetric and, the engineer must
confirm, positive definite before treating it as a covariance. The
leading minors are $4 > 0$, $4\cdot5 - 2^{2} = 16 > 0$, and $\det
\mat{C} = 4(20-4) - 2(8-0) = 64 - 16 = 48 > 0$, so Sylvester's
criterion passes and the matrix is a legitimate covariance.

An estimator downstream (a least-squares fit, a fault detector) assumes
independent unit-variance noise on each channel. The step that makes
that assumption true is whitening: transform the raw residual
$\vect{r}$ by $\mat{C}^{-1/2}$ so that $\vect{z} = \mat{C}^{-1/2}
\vect{r}$ has covariance $\mat{C}^{-1/2}\mat{C}\mat{C}^{-1/2} =
\mat{I}$. Computing $\mat{C}^{-1/2}$ is a matrix-function problem
solved through the spectrum. Diagonalise $\mat{C} = \mat{Q}\mat{\Lambda}
\mat{Q}\transpose$; the eigenvalues are $\lambda = (1.63,\ 4.00,\ 7.37)$
to three digits, all positive as the definiteness check promised, and
$\mat{C}^{-1/2} = \mat{Q}\diag(\lambda_i^{-1/2})\mat{Q}\transpose$. The
whitening matrix scales each eigen-direction by $1/\sqrt{\lambda_i}$:
the noisiest direction (eigenvalue $7.37$, the common-mode axis where
the two coupling entries reinforce) is shrunk hardest, by a factor
$1/\sqrt{7.37} \approx 0.368$, so a fault detector applied to $\vect{z}$
no longer mistakes the loud shared-noise axis for a signal.

The reverse operation generates correlated test noise from a random
generator that only produces white noise. Draw $\vect{w}$ with
independent unit-variance entries and form $\vect{r} = \mat{C}^{1/2}
\vect{w}$ with $\mat{C}^{1/2} = \mat{Q}\diag(\sqrt{\lambda_i})\mat{Q}
\transpose$; the result has covariance exactly $\mat{C}$, reproducing
the bench-measured correlation structure for a Monte Carlo study of the
estimator. Either direction could use the Cholesky factor $\mat{L}$
with $\mat{L}\mat{L}\transpose = \mat{C}$, which is cheaper, and for
plain sampling the engineer usually does. The symmetric square root
earns its extra cost only where the symmetry is load-bearing: when the
whitening must not rotate the axes (so the whitened residuals stay
interpretable per channel), or when the covariance is interpolated
between two operating points, where the Cholesky factor's dependence on
the arbitrary column order produces a discontinuous path and the
symmetric root does not. The eigendecomposition of section 10.2, the
definiteness test of section 10.3, and the matrix-function identity of
this section combine into one estimation-pipeline step.

\subsection{Worked example: the matrix exponential, numerically}

Consider the linear system $\dot{\vect{x}} = \mat{A}\vect{x}$ with

\[
\mat{A} = \begin{pmatrix} -2 & 1 \\ 0 & -3 \end{pmatrix}.
\]

The matrix is upper triangular, so its eigenvalues are $\lambda_{1} =
-2$ and $\lambda_{2} = -3$, both with eigenvectors found from
$(\mat{A} - \lambda_{i}\mat{I})\vect{v}_{i} = \vect{0}$:
$\vect{v}_{1} = (1, 0)\transpose$ and $\vect{v}_{2} = (1, -1)
\transpose$. Then $\mat{S} = \begin{psmallmatrix} 1 & 1 \\ 0 & -1
\end{psmallmatrix}$, $\mat{S}\inv = \begin{psmallmatrix} 1 & 1 \\ 0 & -1
\end{psmallmatrix}$ (this $\mat{S}$ is its own inverse here), and

\[
e^{\mat{A}t} = \mat{S}\begin{pmatrix} e^{-2t} & 0 \\ 0 & e^{-3t}
\end{pmatrix}\mat{S}\inv = \begin{pmatrix} e^{-2t} & e^{-2t} - e^{-3t}
\\ 0 & e^{-3t} \end{pmatrix}.
\]

The reader can check the two defining properties: $e^{\mat{A}\cdot 0} =
\mat{I}$, and $\frac{d}{dt}e^{\mat{A}t} = \mat{A}e^{\mat{A}t}$. The
$(1,2)$ entry $e^{-2t} - e^{-3t}$ is the cross-coupling: a disturbance
in the second state bleeds into the first through the off-diagonal $1$,
peaks at $t = \ln(3/2) \approx 0.405$, then decays as the slower mode
$e^{-2t}$ wins. This is the same coupling-then-decay shape that the
non-normal transient of section 10.7 amplifies; here, with a modest
off-diagonal, it stays small. The eigenvalues set the asymptotic decay
rates; the eigenvectors, through $\mat{S}$, set how much the modes mix
on the way down. Production code does not compute $e^{\mat{A}t}$ by this
hand route for any matrix that might be near-defective; it uses
scaling-and-squaring with a Pad\'e approximant (the design exercise at
the end of the chapter), which the failure section's reasoning explains.

\subsection{Jordan form and the defective exponential}\pathtag{mastery}

When the matrix is defective, the spectral formula needs the Jordan
form. Take the genuinely defective

\[
\mat{A} = \begin{pmatrix} \lambda & 1 \\ 0 & \lambda \end{pmatrix},
\]

a single Jordan block. Write $\mat{A} = \lambda\mat{I} + \mat{N}$ with
$\mat{N} = \begin{psmallmatrix} 0 & 1 \\ 0 & 0 \end{psmallmatrix}$
nilpotent, $\mat{N}^{2} = \mat{0}$. Because $\lambda\mat{I}$ and
$\mat{N}$ commute, $e^{\mat{A}t} = e^{\lambda t}e^{\mat{N}t}$, and the
series for $e^{\mat{N}t}$ terminates after two terms:

\[
e^{\mat{A}t} = e^{\lambda t}(\mat{I} + t\mat{N}) = e^{\lambda t}
\begin{pmatrix} 1 & t \\ 0 & 1 \end{pmatrix}.
\]

The off-diagonal carries the factor $t$. A defective system's response
is not a pure exponential; it is a polynomial in $t$ times an
exponential, here the term $t\,e^{\lambda t}$. When $\lambda < 0$ the
factor $t$ delays the decay, producing a hump before the exponential
finally wins, and when $\lambda$ is a pure imaginary the linear growth
$t$ is unbounded oscillation rather than the bounded oscillation of a
simple eigenvalue. The engineer who models a system as diagonalisable
and predicts a clean $e^{\lambda t}$ misses the $t\,e^{\lambda t}$ term
entirely at a defect, which is one more reason the parameter-sweep
collision of section 10.7 deserves the caution it gets. The general
$k \times k$ Jordan block produces terms up to $t^{k-1}e^{\lambda t}$,
the polynomial degree set by the size of the block.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-jordan-transient.tex}

\Cref{fig:vol02:ch10:jordan-transient} contrasts the defective response
with the pure exponential a diagonalisable matrix would give at the
same eigenvalue. The extra factor of $t$ pushes the response up to a
hump before the exponential pulls it down, peaking at $t = -1/\lambda$,
a transient a diagonalisable model does not predict. The picture is the
time-domain reading of the algebra: the defect is not a divergence when
$\lambda < 0$, but it is a bulge the naive model misses, and near the
collision the bulge is what tips a linearised system past a saturation
limit before its spectrum's verdict of decay takes hold.

\subsection{Cayley-Hamilton in computation}\pathtag{mastery}

The Cayley-Hamilton theorem of section 10.2 is not only an algebraic
curiosity; it computes. Because $\mat{A}$ satisfies its own degree-$n$
characteristic polynomial, every power $\mat{A}^{k}$ with $k \geq n$ is
a linear combination of $\mat{I}, \mat{A}, \ldots, \mat{A}^{n-1}$, and
so is any matrix function expressible as a power series. For the
$2 \times 2$ matrix $\mat{A} = \begin{psmallmatrix} 4 & 1 \\ 2 & 3
\end{psmallmatrix}$ of section 10.1, with characteristic polynomial
$\lambda^{2} - 7\lambda + 10$, the theorem gives $\mat{A}^{2} =
7\mat{A} - 10\mat{I}$ directly, without forming the product. Higher
powers follow by repeated substitution: $\mat{A}^{3} = 7\mat{A}^{2} -
10\mat{A} = 7(7\mat{A} - 10\mat{I}) - 10\mat{A} = 39\mat{A} -
70\mat{I}$, and so on, every power reduced to the pair $(\mat{A},
\mat{I})$ with scalar coefficients.

The systematic version computes $f(\mat{A})$ for any $f$ by matching on
the spectrum. Seek a polynomial $p(\lambda) = c_{0} + c_{1}\lambda +
\cdots + c_{n-1}\lambda^{n-1}$ of degree $n - 1$ with $p(\lambda_{i}) =
f(\lambda_{i})$ at every eigenvalue (and, at a repeated eigenvalue, with
matching derivatives, $p'(\lambda_{i}) = f'(\lambda_{i})$, and so on up
to the multiplicity). Then $f(\mat{A}) = p(\mat{A})$, because $f$ and
$p$ agree on the spectrum and the matrix function depends only on the
values of $f$ there. For the same $\mat{A}$ with eigenvalues $5$ and
$2$, to compute $e^{\mat{A}}$ solve the two equations $c_{0} + 5c_{1} =
e^{5}$ and $c_{0} + 2c_{1} = e^{2}$, giving $c_{1} = (e^{5} - e^{2})/3$
and $c_{0} = e^{2} - 2c_{1}$, then $e^{\mat{A}} = c_{0}\mat{I} +
c_{1}\mat{A}$. The method is the Krylov-subspace idea in miniature
(Volume II Chapter 16): a matrix's action on the full space is captured
by its first $n$ powers, and a function of the matrix is a degree-$(n-1)$
polynomial whose coefficients are read off the spectrum.

\subsection{Total least squares through the SVD}\pathtag{standard}

Ordinary least squares of Chapter 9 assumes the matrix $\mat{A}$ in
$\mat{A}\vect{x} \approx \vect{b}$ is known exactly and only $\vect{b}$
carries error. When both the inputs and the outputs are measured with
noise, the right model is \engterm{total least squares}, which allows
corrections to $\mat{A}$ as well as to $\vect{b}$ and finds the
smallest joint correction that makes the system consistent.
Geometrically, ordinary least squares minimises vertical distances to a
fitted line; total least squares minimises perpendicular distances, the
honest choice when the horizontal axis is as noisy as the vertical.

The SVD solves it cleanly. Stack the data as the augmented matrix
$[\,\mat{A}\;\;\vect{b}\,]$, compute its SVD, and the total-least-squares
correction is the rank-one Eckart-Young perturbation that removes the
smallest singular value. Concretely, if the augmented matrix has
singular value decomposition with smallest right singular vector
$\vect{v}_{n+1}$, partitioned as $\vect{v}_{n+1} = (\vect{v}_{x},
v_{b})\transpose$ with $\vect{v}_{x}$ of length $n$ and $v_{b}$ a
scalar, then the total-least-squares solution is $\vect{x}_{\text{TLS}}
= -\vect{v}_{x}/v_{b}$, provided $v_{b} \neq 0$. The solution is the
direction the augmented data least supports, normalised back into the
original coordinates. The residual is the smallest singular value of the
augmented matrix, the exact Eckart-Young error of dropping the last
component. When $v_{b} = 0$ the smallest singular direction lies
entirely in the $\mat{A}$ block, the columns of $\mat{A}$ are themselves
nearly dependent, and the total-least-squares problem has no solution: a
genuine signal that the inputs themselves are degenerate, beyond any
trouble in the fit. Total
least squares is the calibration engineer's tool whenever the
independent variable is itself an instrument reading, and the SVD is
what makes it a one-line computation rather than a nonlinear
optimisation.

\begin{example}
A calibration bench fits a proportional model $y = a x$ to four paired
readings $(x, y)$ where both instruments are noisy: $(1, 1.2)$,
$(2, 1.8)$, $(3, 3.4)$, $(4, 3.6)$. Ordinary least squares treats $x$
as exact and minimises the vertical residuals, giving slope $a_{\text{OLS}}
= (\sum x_i y_i)/(\sum x_i^2) = 29.4/30 = 0.980$. Total least squares, honest about
noise in $x$, minimises the perpendicular residuals through the
augmented SVD. Stack the data as $[\,\mat{A}\;\vect{b}\,]$ with $\mat{A}
= (1, 2, 3, 4)\transpose$ and $\vect{b} = (1.2, 1.8, 3.4, 3.6)
\transpose$, a $4 \times 2$ matrix. Its singular values are $\sigma_1 =
7.68$ and $\sigma_2 = 0.444$, and the smallest right singular vector is
$\vect{v}_2 = (-0.702,\ 0.712)\transpose$. Partitioning $\vect{v}_2 =
(v_x, v_b)$ with $v_x = -0.702$ and $v_b = 0.712$, the total-least-squares
slope is $a_{\text{TLS}} = -v_x/v_b = 0.702/0.712 = 0.986$. The TLS
slope is slightly steeper than the OLS slope, the standard consequence
of accounting for the horizontal scatter that OLS pretends away, and
the residual is exactly the smallest singular value $\sigma_2 = 0.444$,
the Eckart-Young cost of collapsing the augmented data onto the
best-fit line. The whole computation is one SVD and a division; the
nonlinear perpendicular-distance minimisation the definition suggests
never has to be run.
\end{example}

\subsection{Krylov methods: Lanczos and Arnoldi}\pathtag{mastery}

Power iteration finds one eigenvector at a time and wastes the
intermediate iterates. The \engterm{Krylov subspace} $\mathcal{K}_{m}
(\mat{A}, \vect{b}) = \spn\{\vect{b}, \mat{A}\vect{b}, \mat{A}^{2}
\vect{b}, \ldots, \mat{A}^{m-1}\vect{b}\}$ keeps all of them, and the
best eigenvalue estimates from an $m$-dimensional Krylov subspace are
far better than the single estimate power iteration extracts after $m$
steps. The \engterm{Arnoldi iteration} builds an orthonormal basis of
$\mathcal{K}_{m}$ by modified Gram-Schmidt, producing
$\mat{A}\mat{Q}_{m} = \mat{Q}_{m}\mat{H}_{m} + \text{(remainder)}$ with
$\mat{H}_{m}$ a small $m \times m$ upper-Hessenberg matrix whose
eigenvalues, the \engterm{Ritz values}, approximate the extreme
eigenvalues of $\mat{A}$. For a symmetric $\mat{A}$ the Hessenberg
matrix is tridiagonal and the Arnoldi recurrence shortens to the
three-term \engterm{Lanczos iteration}, the method of choice for the
lowest or highest few eigenvalues of a large sparse symmetric matrix.

The engineering import is scale. A stiffness matrix from a
finite-element model can have $n$ in the millions, far too large to
diagonalise densely, yet the engineer wants only the lowest dozen
natural frequencies. Lanczos delivers them in a few dozen
matrix-vector products against the sparse $\mat{A}$, never forming a
dense factor, converging fastest at the ends of the spectrum where the
modes of interest live. The cost is one subtlety: in floating point the
Lanczos vectors lose orthogonality once a Ritz value converges,
producing spurious duplicate eigenvalues (\engterm{ghosts}) that
re-orthogonalisation or the standard restarted variants remove. The
judgment exercise at the end of the chapter sizes the Lanczos-versus-
dense choice; the foundational treatment is in
\cite[ch.~9-10]{text:v2ch10-golub-vanloan2013} and the numerical-stability
account in \cite[ch.~36]{text:trefethen-bau1997}.

\subsection{The QR algorithm with shifts}\pathtag{mastery}

Section 10.4 introduced the unshifted QR algorithm as the engine behind
dense eigensolvers. Unshifted, it converges painfully slowly: the
trailing subdiagonal entry shrinks at the ratio
$\abs{\lambda_{n}/\lambda_{n-1}}$ per step, so a clustered spectrum
crawls. The fix that turned QR from a curiosity into the dominant
dense eigenalgorithm is \engterm{shifting}. Before each factor-and-
reverse step, subtract a scalar estimate $\mu_{k}$ of the target
eigenvalue: factor $\mat{A}_{k} - \mu_{k}\mat{I} = \mat{Q}_{k}\mat{R}_{k}$,
form $\mat{A}_{k+1} = \mat{R}_{k}\mat{Q}_{k} + \mu_{k}\mat{I}$. The
shift is still a similarity transform, so the eigenvalues are preserved,
but it accelerates the bottom-right corner toward the eigenvalue
nearest $\mu_{k}$, exactly as the shift accelerates inverse iteration
in section 10.6. The \engterm{Wilkinson shift} takes $\mu_{k}$ as the
eigenvalue of the trailing $2 \times 2$ block closest to the corner
entry, which gives near-cubic convergence of the trailing subdiagonal
for symmetric matrices.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-qr-convergence.tex}

\Cref{fig:vol02:ch10:qr-convergence} contrasts the two rates. Once the
trailing subdiagonal entry is negligible, the corner holds a converged
eigenvalue, the matrix \engterm{deflates} to size $n - 1$, and the
algorithm continues on the leading block. The full production recipe,
the Francis double-shift QR with a preliminary reduction to Hessenberg
form, computes all eigenvalues of a dense $n \times n$ matrix in
$\oom{n^{3}}$ and is the algorithm inside every LAPACK call to
\texttt{?GEEV} or \texttt{?SYEV} \cite[ch.~7-8]{text:v2ch10-golub-vanloan2013}.
The reader who calls an eigensolver is calling shifted QR; the figure
is what its inner loop is doing.

\subsection{A random-matrix view of the noise floor}\pathtag{mastery}

The scree plot of section 10.4 asks the engineer to find the elbow
between structural and noise components by eye. Random matrix theory
turns that judgment into a threshold. If a data matrix is pure noise,
$m$ samples of $n$ independent standardised features, the eigenvalues
of its sample covariance matrix do not pile up at one; they spread
across a band whose edges the \engterm{Marchenko-Pastur law} predicts
exactly. With aspect ratio $\gamma = n/m \leq 1$, the noise eigenvalues
fill the interval $[\lambda_{-}, \lambda_{+}]$ with $\lambda_{\pm} =
(1 \pm \sqrt{\gamma})^{2}$.

\begin{principle}[Marchenko-Pastur noise floor]
For a sample covariance matrix built from $m$ samples of $n$
independent unit-variance features, the eigenvalues attributable to
sampling noise lie, for large $m$ and $n$ with $\gamma = n/m$ fixed,
within $[\,(1 - \sqrt{\gamma})^{2},\, (1 + \sqrt{\gamma})^{2}\,]$. An
observed eigenvalue above $(1 + \sqrt{\gamma})^{2}$ exceeds what noise
alone can produce and marks a genuine principal component; an
eigenvalue inside the band is statistically indistinguishable from
noise.
\end{principle}

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-marchenko-pastur.tex}

\Cref{fig:vol02:ch10:marchenko-pastur} shows the law as a significance
test on the scree plot. The threshold sharpens the chapter's recurring
honesty check: a component whose eigenvalue sits inside the
Marchenko-Pastur bulk is noise the engineer should not retain, however
tempting the cumulative-variance curve makes it. The bridge
structural-health case study below applies the test to real-shaped data
and finds it more conservative than the scree elbow, which is the point:
the elbow
flatters borderline components, and the noise floor calls them out. The
law assumes independent entries and breaks for heavy-tailed or strongly
correlated noise, where the corrected bulk edges of the spiked-covariance
literature replace the plain formula; the engineer who needs the test on
genuinely non-Gaussian data should reach for those corrections rather
than the textbook edges.

\section{Markov chains and stochastic matrices}\pathtag{standard}

A Markov chain is a stochastic process that evolves on a discrete
state space and whose transition probabilities depend only on the
current state. The mathematics is governed by a stochastic matrix
whose spectral properties decide the long-time behaviour. We treat
the discrete-time finite-state case, which is enough for the
PageRank example of section 10.6 and for Volume II Chapter 13's
fuller treatment.

\subsection{Stochastic matrices}

A square matrix $\mat{P} \in \R^{n \times n}$ is \engterm{stochastic}
when its entries are non-negative and each row sums to one:
$P_{ij} \geq 0$ and $\sum_{j} P_{ij} = 1$ for every $i$. The entry
$P_{ij}$ is the probability of transitioning from state $i$ to
state $j$ in one step. The state of the chain at time $t$ is a row
vector $\vect{\pi}_{t} \in \R^{n}$ with non-negative entries summing
to one; the update is $\vect{\pi}_{t+1} = \vect{\pi}_{t} \mat{P}$.

A stochastic matrix has $\lambda = 1$ as an eigenvalue, with the
all-ones column vector $\vect{1}$ as a right eigenvector
($\mat{P} \vect{1} = \vect{1}$). Every other eigenvalue satisfies
$\abs{\lambda} \leq 1$, by the Gerschgorin circle theorem applied
to the row-sum bound \cite[ch.~6]{text:strang2016}. The spectral
radius of a stochastic matrix is exactly $1$.

\subsection{Stationary distribution}

A \engterm{stationary distribution} is a row vector $\vect{\pi}$
with non-negative entries summing to one such that $\vect{\pi}
\mat{P} = \vect{\pi}$. Equivalently, $\vect{\pi}$ is a left
eigenvector of $\mat{P}$ at eigenvalue $1$, normalised to have
entries summing to one. Every finite-state Markov chain has at
least one stationary distribution, by the existence of a left
eigenvector at $\lambda = 1$.

When the chain is \engterm{irreducible} (every state is reachable
from every other state in some number of steps) and
\engterm{aperiodic} (the greatest common divisor of return times to
any state is one), the stationary distribution is unique and the
chain converges to it from every starting distribution. This is the
content of the Perron-Frobenius theorem applied to stochastic
matrices.

\begin{theorem}[Perron-Frobenius, recognition level]
Let $\mat{P}$ be an irreducible aperiodic stochastic matrix. Then
$\lambda = 1$ is a simple eigenvalue, every other eigenvalue
satisfies $\abs{\lambda} < 1$, and there is a unique stationary
distribution $\vect{\pi}^{*}$ with strictly positive entries. The
iterate $\vect{\pi}_{t} \to \vect{\pi}^{*}$ for every starting
distribution $\vect{\pi}_{0}$, with rate of convergence governed by
the second-largest eigenvalue magnitude.
\end{theorem}

We quote the theorem and use it. The full statement covers
non-negative matrices more generally and appears in
\cite[ch.~8]{text:strang2016}. The engineering reading is direct.
The stationary distribution is the long-run state occupancy. The
second-largest eigenvalue magnitude is the mixing rate: a chain
with $\abs{\lambda_{2}} = 0.99$ takes about $1/(1 - 0.99) = 100$
steps to forget initial conditions; a chain with $\abs{\lambda_{2}} =
0.5$ takes about two.

\subsection{Worked example: a two-state chain}

Consider a two-state weather chain with transition matrix

\[
\mat{P} = \begin{pmatrix} 0.7 & 0.3 \\ 0.4 & 0.6 \end{pmatrix},
\]

where state $1$ is sunny and state $2$ is rainy, and the entries
are conditional probabilities of tomorrow given today. The
characteristic polynomial is $\lambda^{2} - 1.3 \lambda + (0.7
\cdot 0.6 - 0.3 \cdot 0.4) = \lambda^{2} - 1.3 \lambda + 0.3$, with
roots $\lambda_{1} = 1$ and $\lambda_{2} = 0.3$.

The stationary distribution comes from the left eigenvector at
$\lambda_{1} = 1$. Solve $\vect{\pi} (\mat{P} - \mat{I}) = 0$:

\[
\vect{\pi} \begin{pmatrix} -0.3 & 0.3 \\ 0.4 & -0.4 \end{pmatrix}
= 0
\quad \Rightarrow \quad
\pi_{1} \cdot (-0.3) + \pi_{2} \cdot 0.4 = 0,
\]

so $\pi_{2}/\pi_{1} = 3/4$. Normalising $\pi_{1} + \pi_{2} = 1$
gives $\vect{\pi}^{*} = (4/7, 3/7) \approx (0.571, 0.429)$. The
chain spends $57.1\%$ of its long-run time sunny and $42.9\%$ rainy.
The second eigenvalue $\lambda_{2} = 0.3$ governs the mixing rate;
about $1/(1 - 0.3) \approx 1.4$ steps suffice to forget initial
conditions, which matches the intuition that the chain has short
memory.

\section{Case study: machine availability as a steady-state problem}\pathtag{standard}
\label{sec:vol02:ch10:markov-casestudy}

The two-state weather chain is a toy. A production reliability
question carries the same eigenvector calculation from a raw
maintenance log to a number on a budget line, and the steps between
are where engineers most often go wrong. We carry one such problem
end to end: a single production machine, modelled as a five-state
Markov chain, whose stationary distribution becomes a long-run
availability and then an annual downtime cost.

\subsection{The problem and the state model}

A factory runs one bottleneck machine across three shifts a day, every
day. Maintenance and production logs over the past year, current as of
2025, are coded into one of five states at the start of each shift:
\emph{running} (full rate), \emph{degraded} (reduced rate, a fault
flagged but the machine still producing), \emph{down} (unplanned
stoppage awaiting diagnosis), \emph{maintenance} (planned or corrective
work in progress), and \emph{spare} (a cold-standby unit swapped in
while the primary is serviced). The question the plant manager asks is
direct. What fraction of shifts will the machine be unavailable in the
long run, and what does that cost per year?

The shift-to-shift transition counts, normalised to probabilities,
give the stochastic matrix

\[
\mat{P} = \begin{pmatrix}
0.90 & 0.07 & 0.02 & 0.01 & 0.00 \\
0.20 & 0.65 & 0.08 & 0.07 & 0.00 \\
0.00 & 0.00 & 0.10 & 0.80 & 0.10 \\
0.50 & 0.10 & 0.00 & 0.20 & 0.20 \\
0.85 & 0.05 & 0.00 & 0.00 & 0.10
\end{pmatrix},
\]

with rows and columns ordered (running, degraded, down, maintenance,
spare). Each row sums to one, the defining property of a stochastic
matrix, and the entries read off the log directly: a running machine
stays running $90\%$ of shifts, slips to degraded $7\%$, fails outright
$2\%$, and is pulled for planned work $1\%$; a down machine almost
always moves to maintenance the next shift ($80\%$), with $10\%$
staying down (diagnosis incomplete) and $10\%$ resolved by swapping in
the spare. The dataset is recorded in
\repopath{data/machine_reliability_chain.csv}.

\subsection{Estimating before solving}

\begin{estimation}
\textbf{Question.} Before computing the stationary distribution,
estimate the long-run fraction of shifts the machine spends running.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The running state is sticky: it returns to itself $90\%$ of the time,
and the largest off-diagonal inflows (the $0.20$ from degraded, the
$0.50$ from maintenance, the $0.85$ from spare) all point back toward
running. A state that retains $90\%$ of its own mass and attracts most
of the rest will dominate the stationary distribution. A first guess
treats the chain as a two-state running-versus-everything-else system:
the machine leaves running at rate $0.10$ per shift and returns from
the lumped other states at a rate around $0.3$ to $0.5$ per shift
(weighting the dominant return entries by how often those states are
occupied). The two-state stationary formula $\pi_{\text{run}} =
\mu_{\text{in}}/(\mu_{\text{in}} + \mu_{\text{out}})$ with
$\mu_{\text{out}} = 0.10$ and $\mu_{\text{in}} \approx 0.28$ gives
$\pi_{\text{run}} \approx 0.28/0.38 \approx 0.74$. The full
calculation returns $0.734$, within a percentage point of the lumped
estimate. The order-of-magnitude habit pays off: the engineer knows
the answer is near three-quarters before forming a single eigenvector,
and a computed value of $0.4$ or $0.95$ would signal a transcription
error in the matrix.
\end{estimation}

\subsection{Solving for the stationary distribution}

The stationary distribution $\vect{\pi}^{*}$ is the left eigenvector
of $\mat{P}$ at eigenvalue $1$, normalised to sum to one:
$\vect{\pi}^{*}\mat{P} = \vect{\pi}^{*}$, equivalently
$\vect{\pi}^{*}(\mat{P} - \mat{I}) = \vect{0}$. The matrix is
irreducible (every state reaches every other: running reaches down in
one step, down reaches running through maintenance, and the spare is
both reachable and exiting) and aperiodic (running has a self-loop, so
the period is one), so by Perron-Frobenius the stationary distribution
is unique and strictly positive, and power iteration from any start
converges to it.

The $5 \times 5$ hand calculation is tedious and error-prone, which is
the point: production code solves it three independent ways and checks
they agree, exactly the discipline the chapter's PCA project imposes.
The reference implementation \repopath{code/machine_reliability.py}
computes $\vect{\pi}^{*}$ by (a) the left eigenvector at eigenvalue
$1$, (b) solving the singular linear system $\vect{\pi}(\mat{P} -
\mat{I}) = \vect{0}$ with the normalisation $\sum_{i}\pi_{i} = 1$
appended as the last equation, and (c) power iteration on the row
vector. All three agree, and the result is

\[
\vect{\pi}^{*} \approx (0.734,\ 0.165,\ 0.031,\ 0.055,\ 0.016),
\]

ordered (running, degraded, down, maintenance, spare). The machine is
running $73.4\%$ of shifts, degraded $16.5\%$, down $3.1\%$, under
maintenance $5.5\%$, and on the spare $1.6\%$.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-markov-convergence.tex}

\Cref{fig:vol02:ch10:markov-convergence} shows the convergence from
the all-running start. The occupancy of the down state climbs from
zero to its stationary $0.031$ within a handful of shifts, and the
slowest few percent of the approach is governed by the second-largest
eigenvalue magnitude $\abs{\lambda_{2}} \approx 0.58$, giving a mixing
time $1/(1 - 0.58) \approx 2.4$ shifts. The chain forgets its starting
condition in under a day of operation, which tells the manager that the
stationary numbers describe the machine's steady behaviour and are not
an artefact of where the analysis window happened to begin.

\subsection{From occupancy to availability to cost}

The engineering reading turns the occupancy vector into a decision
number. A shift counts as \emph{available} (the machine is producing)
when it is running or degraded, so the long-run availability is

\[
A = \pi_{\text{run}} + \pi_{\text{deg}} \approx 0.734 + 0.165 = 0.899,
\]

an availability of $89.9\%$ and an unavailability of $10.1\%$. At three
shifts a day over $365$ days, that is $1095$ shifts a year, of which
$0.101 \times 1095 \approx 111$ shifts are lost to the down and
maintenance and spare-swap states. At an illustrative lost
contribution margin of $4{,}000$ currency units per unavailable shift,
the annual downtime cost is

\[
C = (1 - A) \times 1095 \times 4{,}000 \approx 4.4 \times 10^{5}
\]

currency units a year. The number is the whole point of the exercise:
a maintenance log has become a budget figure through one eigenvector
calculation, and the manager can now compare the cost of the status
quo against the cost of an intervention.

\subsection{The intervention, as a change to the matrix}

Suppose a condition-monitoring upgrade catches faults earlier, so a
degraded machine is twice as likely to be pulled for planned
maintenance before it fails outright. In the matrix this moves mass
within the degraded row: the degraded-to-down entry drops from $0.08$
to $0.04$ and the degraded-to-maintenance entry rises from $0.07$ to
$0.11$, leaving the row sum at one. Re-solving the chain with the
edited row gives a new stationary distribution with a lower down-state
occupancy and a slightly higher maintenance occupancy. Whether the
upgrade pays depends on the sign of the change in $C$: planned
maintenance shifts are still unavailable shifts, so the saving comes
only if catching faults early enough shortens the total time out of
production, not merely relabels it. The matrix makes the trade
explicit, and the engineer evaluates it by recomputing one eigenvector
rather than by argument. This is the recurring shape of a Markov
reliability model: a proposed change is a perturbation of the
transition matrix, and its value is the resulting change in a
stationary functional, computed, not asserted.

\subsection{Where the model fails}

The model assumes the transition probabilities are constant and that
the next state depends only on the current one. Both assumptions break
in ways the engineer must name. A machine that has been degraded for
ten consecutive shifts is more likely to fail than one degraded for
one shift, which is history dependence the Markov property discards;
the fix is to split degraded into ``recently degraded'' and ``long
degraded'' states, trading model size for fidelity. Wear-out makes the
running-to-degraded probability rise over the machine's life, so the
matrix estimated from this year's log will misstate next year's
behaviour; the half-life of the numbers is the maintenance interval,
not the decade. And a transition matrix estimated from a year of
shifts has sampling error in every entry, propagating through the
eigenvector to an error bar on the availability. The Weyl and
Bauer-Fike bounds of section 10.1 size that propagation: the
non-symmetric $\mat{P}$ has a moderately conditioned eigenvector
matrix here, so a $1\%$ error in the entries gives a few-percent error
in $\vect{\pi}^{*}$, and the manager should quote the availability as
$90\% \pm 2\%$, not $89.9\%$.

\section{Power iteration and PageRank}\pathtag{standard}

The dominant eigenvector of a matrix can be computed by repeated
multiplication. Power iteration is the simplest such algorithm and
the conceptual backbone of PageRank, the original Google ranking
algorithm. Production eigensolvers do better, but power iteration
is what every engineer should understand first.

\subsection{Power iteration}

Let $\mat{A}$ be a diagonalisable matrix with eigenvalues $\abs{\lambda_{1}}
> \abs{\lambda_{2}} \geq \cdots \geq \abs{\lambda_{n}}$ and
corresponding eigenvectors $\vect{v}_{1}, \ldots, \vect{v}_{n}$. The
strict inequality at the top is the \engterm{spectral gap}, the
condition that makes power iteration work.

\begin{principle}[Power iteration]
Starting from any vector $\vect{x}_{0}$ with a nonzero component
along $\vect{v}_{1}$, define $\vect{x}_{t+1} = \mat{A} \vect{x}_{t} /
\norm{\mat{A} \vect{x}_{t}}_{2}$. Then $\vect{x}_{t}$ converges to
$\vect{v}_{1}$ (up to sign), and the Rayleigh quotient $\vect{x}_{t}
\transpose \mat{A} \vect{x}_{t}$ converges to $\lambda_{1}$. The
convergence rate is governed by the ratio $\abs{\lambda_{2}/\lambda_{1}}$,
with the error after $t$ steps decaying like $\abs{\lambda_{2}/
\lambda_{1}}^{t}$.
\end{principle}

The proof is one line of algebra in the eigenbasis. Write
$\vect{x}_{0} = \sum_{i} c_{i} \vect{v}_{i}$. Then $\mat{A}^{t}
\vect{x}_{0} = \sum_{i} c_{i} \lambda_{i}^{t} \vect{v}_{i} =
\lambda_{1}^{t} (c_{1} \vect{v}_{1} + \sum_{i \geq 2} c_{i}
(\lambda_{i}/\lambda_{1})^{t} \vect{v}_{i})$. Each ratio inside the
sum has magnitude less than one, so the sum tends to zero, and the
normalised iterate tends to $\vect{v}_{1}$.

The cost per iteration is one matrix-vector product, $\oom{n^{2}}$
flops for a dense matrix and much less for a sparse one. When the
spectral gap is healthy ($\abs{\lambda_{2}/\lambda_{1}} \leq 0.5$,
say), a few dozen iterations suffice. When the gap is small,
convergence drags and a more sophisticated method (Lanczos for
symmetric, Arnoldi for general) is preferred. Production code uses
power iteration only when the gap is known to be large or when the
matrix is too large for a full eigensolver to be feasible.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-power-convergence.tex}

\Cref{fig:vol02:ch10:power-convergence} plots the error against
iteration count for three spectral gaps. The straight lines on the
logarithmic axis are the visible form of geometric convergence: the
slope is $\log\abs{\lambda_{2}/\lambda_{1}}$, so a gap of $0.5$ reaches
three digits in about ten steps and a gap of $0.9$ needs roughly
sixty-six. The plot is the diagnostic an engineer draws when an
eigensolver seems slow; a curved line rather than a straight one
signals that the matrix is not diagonalisable in the assumed way, or
that two eigenvalues of nearly equal magnitude are competing for
dominance.

\subsection{Inverse iteration and shift-and-invert}

Power iteration finds the eigenvalue of largest magnitude. The
engineer usually wants a different one: the smallest eigenvalue of a
stiffness matrix (the fundamental frequency), or the eigenvalue
nearest a known resonance. \engterm{Inverse iteration} delivers it.
Apply power iteration to $(\mat{A} - \mu \mat{I})\inv$ for a chosen
\engterm{shift} $\mu$. The eigenvalues of $(\mat{A} - \mu \mat{I})
\inv$ are $1/(\lambda_{i} - \mu)$, and the one of largest magnitude
corresponds to the $\lambda_{i}$ closest to $\mu$. Power iteration on
the inverse therefore converges to the eigenvector whose eigenvalue is
nearest the shift, at geometric rate set by the ratio of the two
smallest values of $\abs{\lambda_{i} - \mu}$.

The inverse is never formed. Each step solves the linear system
$(\mat{A} - \mu \mat{I})\vect{y}_{t+1} = \vect{x}_{t}$ by a
factorisation of $\mat{A} - \mu \mat{I}$ computed once and reused. The
near-singularity of $\mat{A} - \mu \mat{I}$ when $\mu$ is close to an
eigenvalue is exactly what makes the method converge fast and does not
spoil the answer: the large component the ill-conditioning amplifies
is precisely the eigenvector direction sought, and the normalisation
at each step keeps the iterate bounded.

\begin{example}
For the matrix $\mat{B} = \begin{psmallmatrix} 3 & 1 & 0 \\ -1 & 3 & 0
\\ 0 & 0 & 5 \end{psmallmatrix}$ of section 10.1, with spectrum
$\{3 \pm i,\, 5\}$, take the shift $\mu = 4.5$, closer to the real
eigenvalue $5$ than to the complex pair (which sit at distance
$\abs{3 + i - 4.5} = \sqrt{1.5^{2} + 1} \approx 1.80$ from the shift,
against $\abs{5 - 4.5} = 0.5$ for the real one). Inverse iteration
from $\vect{x}_{0} = (1, 1, 1)\transpose/\sqrt{3}$ gives Rayleigh
quotients $4.733$, $4.977$, $4.998$, $4.99986$, $4.9999999$ over five
steps, converging to $\lambda = 5$ at the geometric rate
$(0.5/1.80)^{t} \approx 0.28^{t}$. The reference implementation in
\repopath{code/inverse_iteration.py} reproduces the sequence and lets
the reader move the shift toward the complex pair to target a
different eigenvalue.
\end{example}

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-inverse-iteration-target.tex}

\Cref{fig:vol02:ch10:inverse-iteration-target} is the picture behind
the method. The spectral transformation $\lambda \mapsto 1/(\lambda -
\mu)$ has a pole at the shift, so the eigenvalue nearest the shift
receives the largest-magnitude image and becomes the target that power
iteration on the shifted inverse converges to. Sliding the shift picks
a different eigenvalue, which is how the engineer targets the mode
nearest a known resonance rather than merely the dominant one.

\begin{example}
The geometric rate of fixed-shift inverse iteration is the ratio of
the two largest transformed magnitudes, and the shift controls it
directly. Take the symmetric spectrum $\{1, 3, 5, 7\}$ and a shift
$\mu = 4.6$ aimed at the eigenvalue $5$. The transformed magnitudes are
$1/\abs{\lambda_i - \mu} = (0.278,\ 0.625,\ 2.50,\ 0.417)$ for the
eigenvalues $(1, 3, 5, 7)$, so the dominant transformed value is
$2.50$ at $\lambda = 5$ and the next largest is $0.625$ at $\lambda =
3$. Inverse iteration therefore converges to the eigenvector at
$\lambda = 5$ at rate $0.625/2.50 = 0.25$ per step, a quarter of the
error removed each iteration, three digits in about five steps. Move
the shift closer, to $\mu = 4.9$: the target magnitude jumps to
$1/0.1 = 10$ while the runner-up at $\lambda = 3$ rises only to
$1/1.9 = 0.526$, so the rate sharpens to $0.526/10 \approx 0.053$ and
three digits arrive in three steps. A shift landing nearer the wanted
eigenvalue buys faster convergence, which is exactly the incentive
that Rayleigh-quotient iteration below exploits by updating the shift
toward the eigenvalue at every step.
\end{example}

\subsection{Rayleigh-quotient iteration}\pathtag{mastery}

Inverse iteration with a fixed shift converges geometrically.
Updating the shift at every step to the current Rayleigh quotient
turns geometric convergence into cubic convergence for a symmetric
matrix. \engterm{Rayleigh-quotient iteration} sets $\mu_{t} =
\vect{x}_{t}\transpose \mat{A} \vect{x}_{t}$, solves
$(\mat{A} - \mu_{t} \mat{I})\vect{y}_{t+1} = \vect{x}_{t}$, and
normalises. The quadratic accuracy of the Rayleigh quotient (section
10.1) feeds back into the shift, and each step roughly triples the
number of correct digits once the iterate is near an eigenvector.

\begin{example}
On the symmetric tridiagonal $\mat{A} = \begin{psmallmatrix} 2 & 1 & 0
\\ 1 & 3 & 1 \\ 0 & 1 & 2 \end{psmallmatrix}$ with spectrum
$\{1, 2, 4\}$, Rayleigh-quotient iteration from $\vect{x}_{0} =
(1, 0.4, -0.2)\transpose$ (normalised) produces shifts $2.667$,
$2.245$, $2.00531$, $2.0000000282$, then the exact $2$. The error
falls from $0.67$ to $0.25$ to $5 \times 10^{-3}$ to $3 \times
10^{-8}$ to machine zero, the digit count tripling each step. The
price of the cubic rate is a fresh factorisation of $\mat{A} - \mu_{t}
\mat{I}$ at every iteration, $\oom{n^{3}}$ per step for a dense
matrix against the $\oom{n^{2}}$ of fixed-shift inverse iteration, so
the method pays only when very high accuracy on one eigenpair is
wanted and the matrix is small or already in Hessenberg or
tridiagonal form. A guard is needed: if a shift lands exactly on an
eigenvalue, $\mat{A} - \mu_{t} \mat{I}$ is singular and the solve
fails; the standard remedy perturbs the shift by a tiny amount and
continues. The cubic convergence is the engine inside the symmetric
QR algorithm's inner loop.
\end{example}

\subsection{The PageRank model}

PageRank, introduced by Page and Brin in 1998 and the spine of
Google's original ranking, is the stationary distribution of a
random walk on the web graph. The web is modelled as a graph with
$n$ pages as nodes and a directed edge $i \to j$ whenever page $i$
links to page $j$. The ``random surfer'' is at page $i$ at time $t$
and moves to one of the pages $i$ links to, chosen uniformly at
random, at time $t+1$.

The transition matrix $\mat{H}$ has entries $H_{ij} = 1/\deg(i)$ if
$i$ links to $j$, zero otherwise, where $\deg(i)$ is the
out-degree of page $i$. Two complications: pages with no outgoing
links (\engterm{dangling nodes}) make $\mat{H}$ not stochastic, and
the web graph is not irreducible (some pages are unreachable from
others). The PageRank model fixes both with a \engterm{teleportation}
parameter $\alpha \in (0, 1)$, typically $\alpha = 0.85$, and a
uniform distribution $\vect{u} = \vect{1}/n$. The corrected matrix
is

\[
\mat{G} = \alpha \mat{H}' + (1 - \alpha) \vect{1} \vect{u}\transpose,
\]

where $\mat{H}'$ replaces dangling rows with $\vect{u}\transpose$ to
make every row sum to one. The matrix $\mat{G}$ is stochastic,
irreducible, and aperiodic by construction. By Perron-Frobenius it
has a unique stationary distribution.

The stationary distribution is the \engterm{PageRank vector}. Each
entry $\pi_{j}$ is the long-run probability that a random surfer is
at page $j$, and pages are ranked by this probability. The original
algorithm computes the PageRank by power iteration on $\mat{G}$.
With $\alpha = 0.85$, the second eigenvalue is bounded by $\alpha$
in absolute value, so convergence is geometric at rate $0.85^{t}$
and about $50$ iterations bring $0.85^{50} \approx 3 \times 10^{-4}$,
adequate for ranking. The matrix $\mat{H}'$ is sparse (most pages
link to few others), so each iteration costs $\oom{m}$ flops, where
$m$ is the number of edges. PageRank, in the original 1998 deployment,
ranked the entire indexed web ($n \approx 10^{8}$ pages, $m
\approx 10^{9}$ edges) in a few hours per run on commodity hardware.

\subsection{The teleportation parameter as engineering choice}

The choice $\alpha = 0.85$ is empirical. Smaller $\alpha$ accelerates
convergence (the spectral gap widens) but makes the ranking depend
more on the uniform distribution and less on the link structure.
Larger $\alpha$ relies more on the links but converges more slowly.
The 0.85 figure was the value the founders settled on for the early
indexes and remains the default in the literature, current as of
2025, when the algorithm is studied as a baseline. Production search
engines have moved to deeper ranking models. PageRank survives as a
worked example of how a spectral algorithm with a stochastic matrix
solves a problem that would be intractable as a combinatorial
optimisation.

\subsection{Worked example: PageRank on a four-page web}

A graph small enough to compute by hand makes the mechanism
concrete. Take four pages with links: page $1 \to 2$ and $1 \to 3$;
page $2 \to 3$; page $3 \to 1$; page $4 \to 3$. Page $1$ has
out-degree two, pages $2$, $3$, $4$ have out-degree one. The raw
link matrix, with $H_{ij} = 1/\deg(i)$ when $i$ links to $j$, is

\[
\mat{H} = \begin{pmatrix}
0 & 1/2 & 1/2 & 0 \\
0 & 0 & 1 & 0 \\
1 & 0 & 0 & 0 \\
0 & 0 & 1 & 0
\end{pmatrix}.
\]

Every row sums to one, so there are no dangling nodes here and the
dangling correction is not needed. The graph is, however, not
strongly connected (no page links to page $4$), so without
teleportation the random surfer can never reach page $4$ and the
stationary distribution would assign it zero, an artefact of the
crawl rather than a judgment of importance. Teleportation fixes it.
With $\alpha = 0.85$ and uniform $\vect{u} = (1/4, 1/4, 1/4, 1/4)$,
the Google matrix is $\mat{G} = 0.85\,\mat{H} + 0.15\,\vect{1}
\vect{u}\transpose$, every entry of which is at least $0.15/4 =
0.0375$, so $\mat{G}$ is strictly positive, irreducible, and
aperiodic. Perron-Frobenius guarantees a unique positive stationary
distribution.

Power iteration from the uniform start $\vect{\pi}_{0} = (1/4, 1/4,
1/4, 1/4)$ updates $\vect{\pi}_{t+1} = \vect{\pi}_{t} \mat{G}$.
After a handful of steps the iterate settles to approximately
$\vect{\pi}^{*} \approx (0.373, 0.196, 0.394, 0.038)$. The ranking
is page $3 > $ page $1 > $ page $2 > $ page $4$. Pages $1$ and $3$
dominate, page $3$ narrowly, because page $3$ is pointed at by three
of the four pages while page $1$ receives the return link from page
$3$ and so recirculates the bulk of the mass; page $4$ scores lowest
because nothing links to it and its only mass comes from
teleportation, $0.15/4 \approx 0.0375$. The numbers match the
intuition that importance flows along
links, with teleportation supplying a small floor so that no page is
exactly zero and no rank-sink can trap all the mass. The reference
implementation in \texttt{code/pagerank\_small.py} reproduces these
values and lets the reader vary $\alpha$ and watch the ranking shift
toward the uniform distribution as $\alpha \to 0$ and toward the
pure link structure as $\alpha \to 1$.

\section{Failure: when the matrix has no real eigenvectors and the engineer panics}\pathtag{core}
\label{sec:vol02:ch10:failure}

Two failure modes complicate eigenvalue analysis for real matrices.
The matrix may be defective, lacking enough eigenvectors to span
the space; or it may have complex eigenvalues, which a real
eigenvector cannot represent. Both are routine in engineering
practice. The discipline is to recognise the regime, choose the
appropriate apparatus, and refuse the temptation to interpret a
defective or complex spectrum as if it were a clean diagonal one.

\subsection{Defective matrices in parameter sweeps}

Defective matrices are not pathological. They are the boundary
between two regimes. Consider the parameter-dependent family

\[
\mat{A}(\epsilon) = \begin{pmatrix} 1 & 1 \\ \epsilon & 1
\end{pmatrix}.
\]

The eigenvalues are $\lambda_{\pm} = 1 \pm \sqrt{\epsilon}$. For
$\epsilon > 0$ they are real and distinct; for $\epsilon < 0$ they
are complex conjugates; for $\epsilon = 0$ they collide at
$\lambda = 1$ and the matrix becomes the Jordan block. The collision
point is where the spectrum changes character. The matrix is
defective there and only there, but a sweep that crosses
$\epsilon = 0$ encounters it generically.

The numerical signature is that the eigenvectors at small $\abs{\epsilon}$
become nearly parallel. The matrix $\mat{S}$ that diagonalises
$\mat{A}(\epsilon)$ has condition number that diverges as
$1/\sqrt{\abs{\epsilon}}$. Computing $\mat{A}^{k}$ via diagonalisation
near the collision amplifies floating-point error by $\kappa(\mat{S})$,
which is exactly when diagonalisation should not be the chosen
algorithm. Production code uses the \engterm{Schur decomposition}
$\mat{A} = \mat{Q} \mat{T} \mat{Q}\transpose$ with $\mat{Q}$
orthogonal and $\mat{T}$ upper triangular: it exists for every
square matrix, is numerically stable, and reads the eigenvalues
off the diagonal of $\mat{T}$.

\subsection{Complex eigenvalues for real matrices}

A real $2 \times 2$ matrix with complex eigenvalues represents a
rotation combined with a scaling. The matrix

\[
\mat{R} = \begin{pmatrix} \cos\theta & -\sin\theta \\ \sin\theta &
\cos\theta \end{pmatrix}
\]

has trace $2\cos\theta$ and determinant $1$; its characteristic
polynomial is $\lambda^{2} - 2\cos\theta \cdot \lambda + 1$, with
roots $\lambda = \cos\theta \pm i \sin\theta = e^{\pm i \theta}$.
For $\theta$ not a multiple of $\pi$, the roots are not real, and
the matrix has no real eigenvectors at all. The action is rotation
through angle $\theta$, which leaves no real direction fixed.

The engineering reading: the matrix is healthy. It just does not
admit a real diagonalisation. Two alternatives exist. Diagonalise
over $\C$ (the eigenvectors are complex, the eigenvalues come in
conjugate pairs, the change-of-basis matrix is complex). Or use the
real Schur form, which produces a block upper-triangular real
matrix with $1 \times 1$ blocks at real eigenvalues and $2 \times 2$
blocks at conjugate-pair eigenvalues. The $2 \times 2$ block carries
the rotation-and-scaling structure intact. Most production solvers
return the real Schur form by default for real input.

\subsection{The engineer's panic, and what to do instead}

A working diagnostic when an engineer encounters a square matrix
whose eigendecomposition appears wrong. First, verify whether
$\mat{A}$ is normal: compute $\mat{A}\transpose \mat{A} -
\mat{A}\mat{A}\transpose$. If non-zero, the matrix is non-normal and
its eigenvectors are not orthogonal; the change-of-basis matrix may
be poorly conditioned. Second, check the condition number of the
eigenvector matrix $\mat{S}$. If $\kappa(\mat{S}) > 10^{8}$, the
matrix is near-defective and diagonalisation amplifies numerical
error; switch to Schur. Third, check whether the eigenvalues are
real. If not, switch to the real Schur form or work over $\C$
explicitly. Fourth, when the application is a matrix function
$f(\mat{A})$ (powers, exponentials, square roots), use a function
algorithm built on Schur (Higham's algorithms in
\cite[ch.~28]{text:trefethen-bau1997}) rather than naively
diagonalising and applying $f$ to the eigenvalues.

\begin{principle}[Refuse naive diagonalisation when the spectrum is
non-trivial]
Diagonalisation is a special-case algorithm. It is the right tool
when the matrix is symmetric, normal, or has well-separated real
eigenvalues. It is the wrong tool when the matrix is defective, has
complex eigenvalues that the application does not handle, or has an
eigenvector matrix whose conditioning is worse than the original
problem. The Schur decomposition exists for every square matrix,
is numerically stable, and is the production default. The engineer
who reaches for diagonalisation should first check that
diagonalisation is what the matrix admits, and refuse the
algorithm when the spectrum says otherwise.
\end{principle}

\subsection{The Schur decomposition as production default}\pathtag{standard}

The Schur decomposition deserves a slower look because every
production eigensolver and matrix-function routine builds on it. A
real square matrix $\mat{A}$ admits $\mat{A} = \mat{Q} \mat{T}
\mat{Q}\transpose$ with $\mat{Q}$ real orthogonal and $\mat{T}$
quasi-upper-triangular: $1 \times 1$ blocks on the diagonal for real
eigenvalues, $2 \times 2$ blocks for complex conjugate pairs. The
eigenvalues read off the diagonal blocks directly. The factorisation
is unique up to ordering of the diagonal blocks and choice of basis
within each block; the LAPACK routine \texttt{?GEES} returns one
canonical choice and offers an interface for reordering the diagonal
\cite[ch.~7]{text:v2ch10-golub-vanloan2013}.

Three properties make Schur the algorithm of choice in regimes where
diagonalisation fails. First, it exists for every square matrix,
including defective ones: there is no Jordan-like degeneracy to
detect at runtime. Second, $\mat{Q}$ is orthogonal, so $\kappa_{2}
(\mat{Q}) = 1$ and the change-of-basis introduces no
error-amplification. Third, $\mat{T}$ shares the eigenvalues of
$\mat{A}$ exactly, so the eigenvalue computation reduces to reading
the diagonal blocks of a numerically reliable triangular matrix.

For matrix functions $f(\mat{A})$ the Schur form leads to the
Parlett recurrence: write $f(\mat{T})$ as an upper triangular matrix
with $f$ of the diagonal blocks on its diagonal and recursively
compute the strictly upper triangular entries from the
diagonal-block ratios. Higham's book on matrix functions is the
production reference \cite[ch.~9]{text:v2ch10-higham-functions2008};
the engineer should reach for \texttt{expm}, \texttt{logm},
\texttt{sqrtm}, and similar library routines rather than coding
$f(\mat{A})$ by hand even when the matrix appears benign, because
the underlying algorithms handle defective and near-defective cases
through Schur and produce useful answers where naive diagonalisation
would produce noise.

\subsection{Pseudospectra and operator non-normality}\pathtag{mastery}

A non-normal matrix has eigenvectors that are not orthogonal. The
eigenvalues then under-describe the matrix's transient behaviour.
The \engterm{pseudospectrum} of $\mat{A}$ at level $\epsilon$ is the
set of complex numbers $z$ such that $\norm{(\mat{A} - z\mat{I})\inv}
\geq 1/\epsilon$. For normal matrices the pseudospectrum reduces to
the union of $\epsilon$-discs around the eigenvalues. For non-normal
matrices the pseudospectrum can extend far from the spectrum,
explaining why a matrix with eigenvalues well inside the unit disc
can still produce large transient growth $\norm{\mat{A}^{k}
\vect{x}_{0}}$ before settling to its asymptotic decay. The
hydrodynamic-stability literature of the 1990s replaced eigenvalue
analysis with pseudospectral analysis for exactly this reason. We
mention the apparatus and refer the interested reader to
\textcite{text:trefethen-bau1997} for the foundational treatment.
The engineering import for this chapter: if a transient analysis
matters and the matrix is non-normal, the eigenvalues alone are not
the right diagnostic and a pseudospectral plot is the next tool.

\begin{mastery}
A cheaper non-normality diagnostic than the full pseudospectrum is the
\engterm{numerical range} (or field of values), the set
$W(\mat{A}) = \{\,\vect{x}^{*}\mat{A}\vect{x} : \norm{\vect{x}}_{2} =
1\,\}$ of all Rayleigh quotients over unit vectors. For a symmetric
matrix the numerical range is exactly the interval $[\lambda_{\min},
\lambda_{\max}]$, the trapping of the Rayleigh-quotient figure above.
For a general matrix the Toeplitz-Hausdorff theorem says $W(\mat{A})$
is a convex region of the complex plane that always contains the
eigenvalues, and for a non-normal matrix it bulges out beyond them. The
gap between the numerical range and the convex hull of the spectrum is
a one-number measure of non-normality: for the normal matrix the two
coincide, and for the transient-growth matrix of the worked example
below the numerical range extends well to the right of the two negative
eigenvalues, its rightmost point being positive even though both
eigenvalues are negative. That positive numerical abscissa is a
guarantee of initial transient growth: the largest initial slope of
$\norm{e^{\mat{A}t}\vect{x}_0}$ equals the rightmost real part in
$W(\mat{A})$, so an eigenvalue analysis reporting only negative real
parts can coexist with a numerical range that certifies the response
grows before it decays. The numerical range costs one small
optimisation rather than the grid of resolvent norms a pseudospectrum
needs, and it answers the initial-growth question exactly.
\end{mastery}

\subsection{Worked example: transient growth from a stable spectrum}

The non-normal trap is best seen on a single small matrix. Consider

\[
\mat{A} = \begin{pmatrix} -0.5 & 10 \\ 0 & -0.6 \end{pmatrix}.
\]

It is upper triangular, so its eigenvalues are the diagonal entries
$-0.5$ and $-0.6$, both negative. A naive reading declares the
discrete system $\vect{x}_{t+1} = \mat{A} \vect{x}_{t}$ and the
continuous system $\dot{\vect{x}} = \mat{A} \vect{x}$ stable: every
eigenvalue has negative real part, so $\norm{e^{\mat{A}t}\vect{x}_{0}}
\to 0$. The asymptotic claim is correct. The transient claim is
where the engineer is ambushed. The off-diagonal $10$ makes the
matrix strongly non-normal; the two eigenvectors are nearly
parallel, $\vect{v}_{1} = (1, 0)\transpose$ and $\vect{v}_{2}
\approx (1, -0.01)\transpose$, with an angle between them of well
under one degree.

Start from $\vect{x}_{0} = (0, 1)\transpose$, a unit vector along
the second coordinate. Write it in the eigenbasis: because the
eigenvectors are nearly parallel, the coefficients $c_{1}$ and
$c_{2}$ are large and nearly cancelling, of magnitude around $100$
each. The continuous solution $e^{\mat{A}t}\vect{x}_{0} = c_{1}
e^{-0.5t}\vect{v}_{1} + c_{2}e^{-0.6t}\vect{v}_{2}$ is a difference
of two large terms that decay at slightly different rates. For early
$t$ the cancellation is incomplete and the norm climbs to roughly
$\norm{e^{\mat{A}t}\vect{x}_{0}}_{\max} \approx 6.7$ near $t \approx
1.8$, a near-sevenfold transient amplification, before the slower
asymptotic decay finally wins. A system whose spectrum says
``decaying'' produces, on the way down, a transient large enough to
trip a saturation limit, exceed a stress allowable, or push a
linearised model outside its valid range. The eigenvalues are not
wrong; they are answering the asymptotic question, and the transient
question has a different answer that only the non-normality, or the
pseudospectrum, or a direct simulation, reveals. The control-system
failure-analysis exercise at the end of the chapter is built on
exactly this trap.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-transient-growth.tex}

\Cref{fig:vol02:ch10:transient-growth} plots the response norm
against time for this matrix and, for contrast, for a normal matrix
with the same two eigenvalues. The normal matrix decays from the first
instant; the non-normal one overshoots to nearly seven times its
starting size before the spectrum's verdict finally takes hold. The
gap between the two curves is the transient the eigenvalues do not
see, and its size is set by the non-orthogonality of the
eigenvectors, which the pseudospectrum quantifies and the spectrum
alone cannot.

\subsection{Real-valued failure: linear stability}

In dynamical systems, eigenvalues of the Jacobian at a fixed point
classify the local behaviour. Real positive eigenvalues mean
exponential growth; real negative ones mean exponential decay;
complex pairs with positive real part mean growing oscillation;
complex pairs with negative real part mean decaying oscillation;
complex pairs on the imaginary axis mean neutral oscillation that
the linearisation cannot resolve. Volume III Chapter 7 develops the
classification in full; the eigenvalue apparatus of this chapter is
what makes the classification computable. A control engineer who
designs a feedback law without checking the closed-loop eigenvalues
has not done the design.

The classification carries a boundary the engineer must respect. For a
continuous-time system $\dot{\vect{x}} = \mat{A}\vect{x}$ the stability
frontier is the imaginary axis: every eigenvalue with negative real
part decays, and the design margin is the distance of the rightmost
eigenvalue from that axis, the \engterm{spectral abscissa}. For a
discrete-time system $\vect{x}_{t+1} = \mat{A}\vect{x}_t$ the frontier
is the unit circle instead, because the response scales by
$\lambda^{t}$ and decays exactly when $\abs{\lambda} < 1$; the margin
is one minus the largest eigenvalue magnitude, the \engterm{spectral
radius}. The two frontiers are joined by the exponential map: a
continuous-time eigenvalue $\lambda$ becomes the discrete-time
eigenvalue $e^{\lambda h}$ under sampling at interval $h$, and the left
half-plane maps to the inside of the unit disc. An engineer who runs a
continuous-time stability test on a system that is actually sampled at
a coarse rate can pass the wrong test, which is the discretisation
failure the closing failure-analysis exercise dissects. Two cautions
from earlier in the chapter attach here. A pair on the imaginary axis
(or on the unit circle) is the neutral case the linearisation cannot
resolve; the nonlinear terms Volume III Chapter 7 carries decide the
verdict, and the eigenvalue analysis honestly reports that it cannot.
And a spectrum comfortably inside the stable region does not by itself
rule out the transient growth of the non-normal worked example above:
the eigenvalues govern the asymptotic fate, and a saturation limit
crossed during a transient can end the system before the asymptote
arrives. Stability analysis is eigenvalue location plus a check that
the transient the eigenvalues do not see stays within bounds.

\section{Worked examples: PCA, modal analysis, image compression}\pathtag{standard}

The chapter has built three apparatus: the eigendecomposition of
square matrices, the SVD of arbitrary matrices, and the spectral
reading of stochastic matrices. Three worked examples consolidate
them on engineering data.

\subsection{Worked PCA on a small dataset}

Consider four samples with two features each, arranged in the
data matrix

\[
\mat{X} = \begin{pmatrix} 2 & 0 \\ 0 & 2 \\ 4 & 4 \\ 6 & 6
\end{pmatrix}.
\]

The column means are $\bar{\vect{x}} = (3, 3)$. Centring gives

\[
\mat{X}_{c} = \begin{pmatrix} -1 & -3 \\ -3 & -1 \\ 1 & 1 \\ 3 & 3
\end{pmatrix}.
\]

The covariance matrix is

\[
\mat{C} = \frac{1}{m-1} \mat{X}_{c}\transpose \mat{X}_{c}
= \frac{1}{3} \begin{pmatrix} 20 & 16 \\ 16 & 20 \end{pmatrix}
= \begin{pmatrix} 20/3 & 16/3 \\ 16/3 & 20/3 \end{pmatrix}.
\]

The eigenvalues of $\mat{C}$ solve $\lambda^{2} - (40/3)\lambda
+ (400/9 - 256/9) = 0$, giving $\lambda^{2} - (40/3)\lambda + 16 =
0$, with roots $\lambda_{1} = 12$ and $\lambda_{2} = 4/3$. The
explained-variance ratio is $12/(12 + 4/3) = 9/10 = 0.9$. The first
principal component captures $90\%$ of the variance.

The principal directions are the eigenvectors of $\mat{C}$. At
$\lambda_{1} = 12$, $(\mat{C} - 12 \mat{I})$ has rows $(-16/3, 16/3)$
and $(16/3, -16/3)$, both giving $v_{1} = v_{2}$. The unit
eigenvector is $\vect{v}_{1} = (1, 1)/\sqrt{2}$. At $\lambda_{2} =
4/3$, the eigenvector is $\vect{v}_{2} = (1, -1)/\sqrt{2}$. The
data, viewed in the $(\vect{v}_{1}, \vect{v}_{2})$ basis, has most
of its spread along $\vect{v}_{1}$ (the diagonal of feature space)
and very little along $\vect{v}_{2}$ (the anti-diagonal). PCA
recovers the structure the data already had: it lies near a line.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-pca-axes.tex}

\Cref{fig:vol02:ch10:pca-axes} draws the same idea on a slightly
larger synthetic scatter, with both principal axes overlaid through
the centroid. The longer axis points along the dominant principal
direction, with length proportional to $\sqrt{\sigma_{1}^{2}/(m-1)}$;
the shorter axis is the perpendicular direction. Projecting the
samples onto the longer axis retains most of the variance, and a
downstream classifier or regressor fed only that projection still
resolves the samples. The dataset under
\texttt{data/synth\_pca\_dataset.csv} provides a $20$-sample,
five-feature version with a planted rank-two structure; the project
in section 10.9 runs PCA on this dataset and on a dataset of the
reader's choice, by three independent routes.

\subsection{Modal analysis of a three-mass system}

Three equal masses $m$ are connected in a line by four equal springs
of stiffness $k$ to walls and to each other. The stiffness matrix
is

\[
\mat{K} = k \begin{pmatrix} 2 & -1 & 0 \\ -1 & 2 & -1 \\ 0 & -1 & 2
\end{pmatrix}.
\]

The matrix is symmetric and positive definite, so it has three
real positive eigenvalues and an orthonormal eigenbasis. The
characteristic polynomial of $\mat{K}/k$ is $-\lambda^{3} + 6
\lambda^{2} - 10 \lambda + 4$, with roots $\lambda_{1} = 2 -
\sqrt{2}$, $\lambda_{2} = 2$, $\lambda_{3} = 2 + \sqrt{2}$. The
natural frequencies are $\omega_{i} = \sqrt{\lambda_{i} k/m}$. Their
ratios are $\omega_{1} : \omega_{2} : \omega_{3} \approx 1 : 1.85
: 2.41$, with the slow mode close to the all-in-phase motion and
the fast mode close to the alternating motion of adjacent masses.

The eigenvectors are the \engterm{mode shapes}. They tell the
engineer which displacement pattern oscillates at which frequency.
The first mode is roughly $(1, \sqrt{2}, 1)/2$, with the centre
mass moving more than the outer ones in phase. The second mode is
$(1, 0, -1)/\sqrt{2}$, with the centre mass stationary. The third
mode is $(1, -\sqrt{2}, 1)/2$, with the centre mass opposite to the
outers. A general initial condition decomposes into a sum over
modes; each mode evolves independently at its own frequency. Modal
analysis is the universal apparatus of structural and acoustic
vibration analysis (Volume III Chapter 9), of electrical resonance
(Volume IV), and of any linear system whose response is the sum of
its natural responses.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-vibration-modes.tex}

\Cref{fig:vol02:ch10:vibration-modes} draws the three mode shapes
schematically. The numerical mode shapes and frequencies are recorded
in \texttt{data/three\_mass\_modal.csv} for the reader who wants to
plot a time-domain superposition from a chosen initial condition.

\subsection{Image compression by truncated SVD}

A grayscale image is a matrix of pixel intensities. A
$1024 \times 1024$ image is a $\R^{1024 \times 1024}$ matrix
with entries between $0$ and $255$. The full storage is about
$10^{6}$ bytes. Computing the SVD and truncating to rank $k$ stores
the image as $\mat{U}_{k} \in \R^{1024 \times k}$,
$\diag(\sigma_{1}, \ldots, \sigma_{k})$, and $\mat{V}_{k}\transpose
\in \R^{k \times 1024}$, for a total of about $2{,}049 k$ entries.
At $k = 50$, the storage is about $10^{5}$ entries, a tenfold
compression.

The visual quality at fixed compression ratio depends on how fast
the singular values of the image decay. Photographs of natural
scenes have spectra that decay quickly (the matrix has effective
rank in the few tens to few hundreds for plausible quality), so
truncation at $k = 50$ to $k = 100$ produces a recognisable image.
Detailed images, text, or images with high-frequency content decay
slowly and lose more at the same $k$. The Eckart-Young theorem
quantifies the trade-off exactly: the spectral-norm error is
$\sigma_{k+1}$, and the Frobenius-norm error is the tail energy.
The reader running the truncated-SVD compression sees the
approximation degrade smoothly as $k$ decreases, and recognises a
visual threshold around the rank where $\sigma_{k}$ drops to a few
percent of $\sigma_{1}$.

JPEG and modern image codecs do not use the SVD as their basis; they
use the discrete cosine transform, which is data-independent and
much cheaper at encode time. The truncated SVD is the
data-dependent reference: the optimal rank-$k$ approximation in the
matrix-norm sense, against which the cheaper transforms are judged.

\subsection{Low-rank structure in a ratings matrix}

The same truncated-SVD apparatus that compresses an image recovers
hidden structure in a ratings matrix. Suppose five users rate four
films on a one-to-five scale, with missing entries left for the
moment. The intuition behind collaborative filtering is that the
ratings matrix is approximately low rank: a small number of latent
factors (a film's position on an action-to-drama axis, a user's
taste along the same axis) explains most of the variation, and the
full $5 \times 4$ table is close to a rank-two or rank-three
approximation.

Take the complete matrix

\[
\mat{R} = \begin{pmatrix}
5 & 5 & 1 & 1 \\
4 & 5 & 1 & 2 \\
1 & 1 & 5 & 5 \\
2 & 1 & 4 & 5 \\
1 & 2 & 5 & 4
\end{pmatrix},
\]

where the first two columns are one genre and the last two are
another, and the first two users prefer the first genre while the
last three prefer the second. The SVD of $\mat{R}$ has singular
values that drop sharply after the second: $\sigma_{1} \approx 13.5$,
$\sigma_{2} \approx 7.5$, $\sigma_{3} \approx 1.4$, and $\sigma_{4}
\approx 0.8$. The rank-two truncation $\mat{R}_{2} = \sigma_{1}
\vect{u}_{1}\vect{v}_{1}\transpose + \sigma_{2}\vect{u}_{2}
\vect{v}_{2}\transpose$ reproduces every entry of $\mat{R}$ to within
about half a rating point, and the Eckart-Young Frobenius error
$\sqrt{\sigma_{3}^{2} + \sigma_{4}^{2}}$ is about $10\%$ of the
total energy $\sqrt{\sum_{i}\sigma_{i}^{2}}$, so two latent factors
carry roughly $90\%$ of the table. The first right
singular vector $\vect{v}_{1}$ has all-positive entries: it is the
``overall popularity'' direction. The second, $\vect{v}_{2}$, has
opposite signs on the two genre blocks: it is the genre axis that
separates the two clusters of films. The first left singular vector
groups users by overall generosity; the second separates the two
taste groups. The latent factors that a recommender system learns
are exactly these singular vectors. Filling missing entries by the
rank-$k$ reconstruction is the matrix-completion problem at the
heart of the recommender systems of Volume VII, and the truncated
SVD is its linear-algebra core.

\subsection{Reading a scree plot}

Each of these applications, PCA, modal analysis, image compression,
recommender factorisation, ends at the same diagnostic plot: the
singular values (or their squares, the variances) against their
index. The plot is the \engterm{scree plot}, named for the loose
rock that accumulates at the base of a cliff, because the
characteristic shape is a steep face followed by a shallow talus.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-scree-plot.tex}

\Cref{fig:vol02:ch10:scree-plot} draws the two readings the
engineer takes from a scree plot. The \engterm{elbow}, the index
where the curve bends from steep to shallow, marks the boundary
between structural components (the cliff face) and noise components
(the talus); choosing $k$ at the elbow keeps the structure and
discards the noise. The \engterm{cumulative explained-variance
curve}, the running sum of $\sigma_{i}^{2}$ normalised by the total,
rises toward one and is read by drawing a horizontal line at the
chosen retention level ($90\%$, $95\%$, $99\%$) and dropping to the
axis. The two readings usually disagree by a few components, and the
chapter's discipline is to report both, then let the downstream task
break the tie. A plot with no clear elbow (the slow power-law decay
of natural images) warns that no honest low-rank truncation exists,
and the engineer who forces one is discarding signal.

\section{Case study: structural-health PCA on a bridge}\pathtag{standard}
\label{sec:vol02:ch10:bridge-casestudy}

The small worked PCA of section 10.8 had two features and four samples,
enough to see the mechanism and too small to see the engineering. A real
structural-health-monitoring problem carries the same SVD from a raw
sensor log to a maintenance decision, and the steps in between, the
conditioning choice, the loadings interpretation, the noise floor, are
where the analysis is won or lost. We carry one such problem end to end:
eight strain gauges on a road bridge, sampled hourly, reduced by PCA to
the two physical modes that explain the deck's response, with the
limitations named.

\subsection{The instrument and the data matrix}

A two-span road bridge carries eight strain gauges, four bonded to the
north span (labelled \texttt{N1}--\texttt{N4}) and four to the south
span (\texttt{S1}--\texttt{S4}), each reporting microstrain once an
hour. Over four days, current as of 2025, the campaign records $m = 96$
hourly readings of $n = 8$ gauges, the data matrix $\mat{X} \in
\R^{96 \times 8}$ stored in \repopath{data/bridge_strain_gauges.csv}.
Each row is one hour; each column is one gauge. The raw readings sit
near $120$ microstrain, a baseline set by the dead load of the deck and
the bonding pre-strain of each gauge, with hourly swings of a few tens
of microstrain driven by temperature and traffic. The question the
bridge owner asks is whether the eight noisy channels can be reduced to
a few interpretable signals that a monitoring dashboard can track and
alarm on, and what those signals physically mean.

\subsection{Centring, and the conditioning choice}

PCA begins with centring, subtracting each column's mean so that the
analysis sees variation rather than baseline. The centred matrix
$\mat{X}_{c} = \mat{X} - \vect{1}\bar{\vect{x}}\transpose$ has zero
column mean to machine precision, the first invariant to check before
anything else. Forgetting to centre is the most common PCA error in the
field: the first principal direction then points at the centroid, near
$120$ microstrain along the all-ones direction, and reports a spurious
$99.9\%$ of variance that is nothing but the offset.

The conditioning choice comes next, and it is the chapter's recurring
discipline made concrete. The covariance route forms $\mat{C} =
\mat{X}_{c}\transpose\mat{X}_{c}/(m-1)$ and diagonalises it; the SVD
route factors $\mat{X}_{c}$ directly. For this dataset $\kappa(\mat{X}_{c})
\approx 67.8$, a benign number, while $\kappa(\mat{C}) =
\kappa(\mat{X}_{c})^{2} \approx 4.6 \times 10^{3}$, the squaring that
Chapter 9 warned about. At this conditioning both routes return the same
answer to many digits, so the squaring does no visible harm here, but
the habit of running the SVD route is what protects the analysis when a
near-collinear pair of gauges drives $\kappa(\mat{X}_{c})$ to $10^{6}$
and the squared $10^{12}$ destroys the small components. The reference
implementation \repopath{code/bridge_pca.py} computes the PCA both ways
and confirms they agree, exactly the cross-check discipline the
chapter's project imposes.

\subsection{Estimating before solving}

\begin{estimation}
\textbf{Question.} Before computing the SVD, estimate the fraction of
the total variance the first principal component will capture, and the
number of components a monitoring dashboard will need.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
Eight gauges on one deck see mostly the same thing: when a truck crosses
or the sun warms the structure, every gauge strains together. That
shared response is one dominant direction, and it should swamp the
rest. A reasonable guess puts the first component above $90\%$ of the
variance. The remaining structure is the difference between the two
spans, the part of the signal where the north gauges and south gauges
disagree, which is real but small compared with the whole-deck response;
call it most of the rest. Everything below that is sensor noise at the
microstrain level. The estimate, then, is one large component near
$90\%$, one modest component carrying most of the remainder, and a noise
tail: a two-component dashboard. The full calculation returns $91.6\%$
and $8.2\%$ for the first two, cumulative $99.8\%$, with the rest below
$0.1\%$ each, confirming the two-component reading and the order of
magnitude of the split.
\end{estimation}

\subsection{The spectrum and the scree reading}

The SVD of $\mat{X}_{c}$ gives singular values $\sigma_{1} \approx 625$,
$\sigma_{2} \approx 187$, then $\sigma_{3} \approx 16$ and a slow tail
down to $\sigma_{8} \approx 9$. The variances $\sigma_{i}^{2}/(m-1)$ are
$4115$, $369$, then $2.6, 1.9, 1.5, 1.4, 1.2, 0.9$. The explained-variance
ratios are

\[
(0.916,\ 0.082,\ 0.0006,\ 0.0004,\ 0.0003,\ 0.0003,\ 0.0003,\ 0.0002),
\]

cumulative $0.916$, $0.998$ at two components. The scree plot shows a
cliff of two and then a flat talus six wide, the textbook
low-rank-plus-noise shape: the elbow is unambiguous at $k = 2$. Two
principal components carry $99.8\%$ of the variation, and the six
discarded components together hold $0.2\%$, the floor of sensor noise.

\subsection{Reading the loadings}

The numbers that make this a diagnostic rather than a compression are
the loadings, the entries of the first two right singular vectors that
say how each gauge contributes to each component. The first principal
direction $\vect{v}_{1}$ has all eight entries positive and of similar
size, near $0.3$ to $0.4$:

\[
\vect{v}_{1} \approx (0.41, 0.40, 0.41, 0.40,\ 0.30, 0.33, 0.25, 0.30),
\]

ordered \texttt{N1}--\texttt{N4} then \texttt{S1}--\texttt{S4}. Every
gauge loads the same sign on PC1, so PC1 is the whole-structure
response: the combined thermal-expansion and traffic-load mode that
lifts or lowers the entire deck at once. That this mode carries $91.6\%$
of the variance is the quantitative statement that the bridge mostly
moves as one piece.

The second direction $\vect{v}_{2}$ splits the gauges by sign:

\[
\vect{v}_{2} \approx (0.27, 0.32, 0.21, 0.37,\ -0.39, -0.35, -0.49,
-0.37),
\]

the four north gauges positive, the four south gauges negative. PC2 is
the span-asymmetry mode, the part of the response in which one span
strains while the other relaxes, the signature of differential loading
or of a stiffness change in one span relative to the other.
\Cref{fig:vol02:ch10:pca-loadings} draws the eight gauges in the
$(\text{PC1}, \text{PC2})$ loadings plane, where the two spans separate
into two clusters. A monitoring dashboard that tracks the PC1 and PC2
scores is watching the whole-deck response and the span-balance
response, two physically meaningful signals distilled from eight raw
channels, and a drift in the PC2 score over months would flag a
developing asymmetry between the spans before any single gauge crossed
an absolute threshold.

\input{volumes/02-form/ch10-eigenvalues-and-decompositions/figures/fig-pca-loadings.tex}

\subsection{Is the second component real? The noise-floor test}

The scree elbow says keep two components, but a borderline second
component invites a sharper test. The Marchenko-Pastur law of the
random-matrix subsection above supplies one. Standardising the data
(dividing each gauge by its
own standard deviation, so the covariance becomes the correlation
matrix) and reading the correlation eigenvalues against the noise band
gives the honest answer. With $\gamma = n/m = 8/96 \approx 0.083$, the
noise bulk runs to $\lambda_{+} = (1 + \sqrt{0.083})^{2} \approx 1.66$.
The correlation eigenvalues are $7.25$, $0.73$, then a tail near and
below $0.01$. The first eigenvalue, $7.25$, towers far above the noise
floor and is unmistakably real. The second, $0.73$, sits below
$\lambda_{+}$, inside the band.

The plain noise-floor test is therefore more conservative than the scree
elbow: it certifies one component and leaves the span-asymmetry
component formally inside the noise. This is not a contradiction to
resolve by picking the friendlier test; it is the genuine status of a
weak-but-structured signal. The asymmetry mode is physically
interpretable and reproducible across the campaign, which argues for
retaining it; the standardised noise floor says it is close enough to
noise that an automated alarm on its score must be set with care and
validated against known load events, not trusted blind. The disciplined
report states both readings and their disagreement. A practitioner who
quotes only the cumulative $99.8\%$ and declares two clean components
has overclaimed; the noise floor is the check that keeps the second
component honest.

\subsection{From components to a maintenance decision}

The engineering payoff is a dashboard the owner can act on. The PC1
score is the whole-deck strain amplitude, tracked against its seasonal
envelope; an excursion above the envelope flags an overload event for
review. The PC2 score is the span balance; a slow secular drift in PC2,
distinct from its daily thermal cycle, is the early indicator of a
span-specific change, a bearing seizing, a joint cracking, a localised
loss of stiffness, that no single gauge would isolate because each gauge
also carries the dominant PC1 signal. Reducing eight correlated channels
to two physical scores is what turns a wall of noisy strain traces into
two lines a maintenance engineer can read at a glance, and the SVD is
the one computation that finds those two lines optimally under the
Eckart-Young criterion.

\subsection{Where the model fails}

PCA on this data assumes the response is linear and the principal
directions are stationary over the campaign. Both assumptions have
half-lives the engineer must name. PCA finds linear combinations of
gauges; a nonlinear effect, a gauge that saturates at high strain or a
joint that responds differently in tension and compression, is not a
linear combination and will leak across many components rather than
concentrate in one, defeating the two-component summary. The principal
directions themselves drift: a temperature-dependent stiffness means the
loadings estimated in summer differ from those in winter, so the PCA fit
has a seasonal half-life and must be re-estimated, not frozen. And the
whole analysis presumes the gauges are healthy; a debonding gauge
injects a large spurious variance that PCA will happily promote to a
principal component, mistaking instrument failure for structural signal.
The defence is the same discipline the chapter has built throughout:
report the scree and the noise floor together, interpret every retained
loading physically, and treat a component that resists physical reading
as a fault to diagnose rather than a finding to publish. The eigenvalues
and singular values are exact; the interpretation is the engineering, and
it is where the judgment lives.

\section{Project}\pathtag{core}

\begin{project}[Implement Principal Component Analysis from scratch]
The reader implements PCA without using any high-level library
shortcut (no \texttt{sklearn.decomposition.PCA},
\texttt{statsmodels.PCA}, or equivalent), runs it on a real
dataset, and produces a comparative analysis with a 1500-word
reflection.

\textbf{Track}. Simulation with analysis. \textbf{Hazard class}: none.
\textbf{Tools}. Paper, pencil, and a programming environment with
linear-algebra primitives (LU, QR, SVD, eigensolver): Python with
NumPy, MATLAB, Julia, Octave, R, or equivalent.

\textbf{Dataset choice}. The reader chooses a dataset of $m \geq 50$
samples and $n \geq 4$ features, of one of the following kinds: a
Volume I project dataset (e.g.\ a measurement campaign with
multiple correlated readings); a public dataset (UCI archive,
the Iris dataset, the Wine dataset, a country-level statistics
panel from the World Bank); or a synthesised dataset where the
reader has implanted known principal directions in known proportions
(this is a useful debugging dataset and is the recommended starting
point). The reader records the dataset, its source or generation
recipe, and the date.

\textbf{Three implementation routes}. The reader computes the PCA
three ways:

\begin{enumerate}[leftmargin=*]
  \item \engterm{Eigendecomposition of the covariance matrix}.
    Centre the data, form $\mat{C} = \mat{X}_{c}\transpose
    \mat{X}_{c}/(m-1)$, call an eigensolver on $\mat{C}$, sort
    eigenvalues in descending order, sort eigenvectors accordingly.
    Record the condition number of $\mat{X}_{c}$ and of $\mat{C}$.
  \item \engterm{Power iteration}. Centre the data; compute the
    dominant eigenvector of $\mat{C}$ by power iteration with at
    least 100 steps; deflate (subtract the rank-one component) and
    repeat for the next several principal components. Implement
    the deflation by hand. Record the spectral gap at each stage.
  \item \engterm{SVD of the centred data matrix}. Centre the data;
    call the SVD primitive on $\mat{X}_{c}$; the principal
    directions are the columns of $\mat{V}$ and the variances are
    $\sigma_{i}^{2}/(m-1)$. Record the condition number of
    $\mat{X}_{c}$.
\end{enumerate}

\textbf{Reporting}. The reader produces:

\begin{itemize}[leftmargin=*]
  \item A table comparing the three routes' top three principal
    directions (sign-aligned for comparison).
  \item A table of the explained-variance ratios for each route.
  \item A scatter plot of the data projected onto the first two
    principal components, with the sample labels (if categorical)
    shown in colour.
  \item A scree plot of the singular values (or the square roots of
    the eigenvalues), on a log-y axis.
  \item Comparison against a library implementation
    (\texttt{numpy.linalg.svd}, \texttt{sklearn.decomposition.PCA},
    or equivalent) on the same dataset, with agreement reported to
    at least four decimal digits in the explained variances and the
    sign-aligned principal directions.
\end{itemize}

\textbf{Reflection (1500 words)}. The reader writes a 1500-word
narrative addressing:

\begin{itemize}[leftmargin=*]
  \item What the SVD reveals about the dataset that a univariate
    summary (mean and variance per feature) does not. The reader
    names at least one principal direction and interprets it in
    terms of the original features.
  \item The numerical conditioning differences among the three
    routes. Why does the eigendecomposition of $\mat{C} =
    \mat{X}_{c}\transpose \mat{X}_{c}$ have the worst conditioning
    of the three? What is the relation between $\kappa(\mat{X}_{c})$
    and $\kappa(\mat{C})$? Tie back to Chapter 9 section 9.7.
  \item How the explained-variance ratio guides the choice of $k$.
    The reader states a rule used in the project (e.g.\ retain
    components until $90\%$ explained variance is reached) and
    discusses its limits.
  \item What the projection loses. The reader picks a sample, shows
    its original feature vector and its rank-$k$ reconstruction,
    and identifies which features are reconstructed accurately and
    which are not.
  \item Whether PCA was the right tool for the chosen dataset, or
    whether a non-linear method (kernel PCA, manifold learning) would
    have been more appropriate. The reflection takes the linearity
    assumption seriously and does not retroactively justify the
    choice of PCA.
\end{itemize}

\textbf{Deliverable}. The dataset, the code, the three sets of
principal directions and explained variances, the scatter plot, the
scree plot, the library comparison, and the 1500-word reflection.
The general editor's reference solution on a synthesised dataset
with known structure is available online; the reader compares only
after completing their own.

\textbf{Time}. Approximately one hour to choose and clean the
dataset, three to four hours to implement the three routes, one
hour to produce the tables and plots, and three to four hours to
write the reflection. Total approximately eight to ten hours.

\textbf{Phases}. The project resolves in four phases.

\begin{enumerate}[leftmargin=*]
  \item \textit{Setup} (about one hour). Choose the dataset; record
    its source, retrieval date, and the variables. Inspect the data
    for missing entries, units, and obvious outliers. Decide whether
    the columns should be standardised (subtracted mean and divided
    by standard deviation) or only centred; record the choice.
  \item \textit{Implementation} (three to four hours). Implement the
    three routes in the order listed (covariance eigendecomposition,
    power iteration with deflation, SVD of the centred data
    matrix). Verify intermediate quantities (the centred data has
    zero column mean to machine precision; the covariance matrix is
    symmetric; the power iteration converges geometrically).
  \item \textit{Reporting} (one to two hours). Produce the tables
    and plots. Run the library comparison and confirm the agreement
    to at least four decimal digits.
  \item \textit{Reflection} (three to four hours). Write the
    1500-word reflection. Read the chapter's section 10.7 once more
    before judging the conditioning differences.
\end{enumerate}

\textbf{Success criteria}. The project is complete when, for the
chosen dataset, all three of the following are demonstrated.

\begin{itemize}[leftmargin=*]
  \item The three implementation routes return principal directions
    that, after sign alignment, agree to at least four decimal
    digits with each other and with the library reference.
  \item The explained-variance ratios from the three routes agree
    to at least four decimal digits with the library reference, and
    sum to one across the full spectrum.
  \item The reflection answers each of the five prompted questions
    explicitly, with a numerical or graphical reference to the
    project's own outputs (not generic textbook statements).
\end{itemize}

\textbf{Common failure modes and recovery}. Four failure modes
recur. The reader who hits one should apply the listed fix.

\begin{itemize}[leftmargin=*]
  \item \textit{Sign-flip mismatches.} Eigenvectors are determined
    up to sign; library and project may differ by signs. The fix
    is to sign-align before comparing (the
    \texttt{code/pca\_three\_routes.py} helper does this).
  \item \textit{Centring forgotten.} PCA on uncentred data points
    the first principal direction at the centroid rather than along
    the data's spread; explained-variance ratios then exceed unity
    or behave nonsensically. The fix is to verify the centred data
    has zero column mean to machine precision before any further
    computation.
  \item \textit{Power iteration on a defective deflated matrix.}
    Repeated deflation on a non-symmetric matrix can leave a
    near-defective residue; power iteration then converges slowly
    or to a spurious vector. The fix is to use the SVD route for
    the production answer and treat power iteration as a teaching
    artefact.
  \item \textit{Reading explained variance from $\mat{C}$'s
    eigenvalues but reporting it as a percentage of the SVD route's
    sum.} The two sums agree to machine precision when the
    arithmetic is consistent; if they do not, a centring or scaling
    step has been applied unequally across routes. The fix is to
    centre once, in one place, and pass the same centred matrix to
    all three routes.
\end{itemize}
\end{project}

\section{Exercises}

The exercises test calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment. The editor
should attach the solution and rubric file before release.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Compute the eigenvalues and eigenvectors of $\mat{A} =
\begin{psmallmatrix} 3 & 1 \\ 1 & 3 \end{psmallmatrix}$. Verify
that the eigenvectors are orthogonal and that the trace and
determinant equal the sum and product of the eigenvalues.
\end{exercise}

\paragraph{Solution sketch.}
Trace is $6$ and determinant is $9 - 1 = 8$, so the characteristic
polynomial is $\lambda^{2} - 6\lambda + 8 = (\lambda - 4)(\lambda -
2)$. Eigenvalues are $\lambda_{1} = 4$ and $\lambda_{2} = 2$. At
$\lambda_{1} = 4$, $\mat{A} - 4\mat{I} = \begin{psmallmatrix} -1 & 1
\\ 1 & -1 \end{psmallmatrix}$ gives $\vect{v}_{1} = (1, 1)/\sqrt{2}$;
at $\lambda_{2} = 2$, $\mat{A} - 2\mat{I} = \begin{psmallmatrix} 1 &
1 \\ 1 & 1 \end{psmallmatrix}$ gives $\vect{v}_{2} = (1, -1)/\sqrt{2}$.
Inner product $\vect{v}_{1}\transpose \vect{v}_{2} = 0$ confirms
orthogonality, as expected for a symmetric matrix. Trace check:
$4 + 2 = 6 = \tr \mat{A}$. Determinant check: $4 \times 2 = 8 = \det
\mat{A}$.

\begin{exercise}
Compute the diagonalisation $\mat{A} = \mat{S} \mat{\Lambda}
\mat{S}\inv$ of $\mat{A} =
\begin{psmallmatrix} 5 & 4 \\ 1 & 2 \end{psmallmatrix}$. Verify by
multiplying out $\mat{S} \mat{\Lambda} \mat{S}\inv$.
\end{exercise}

\paragraph{Solution sketch.}
Trace is $7$, determinant is $10 - 4 = 6$. Characteristic polynomial
$\lambda^{2} - 7\lambda + 6 = (\lambda - 6)(\lambda - 1)$ gives
$\lambda_{1} = 6$ and $\lambda_{2} = 1$. At $\lambda_{1} = 6$,
$\mat{A} - 6\mat{I} = \begin{psmallmatrix} -1 & 4 \\ 1 & -4
\end{psmallmatrix}$ gives $\vect{v}_{1} = (4, 1)\transpose$ (any
nonzero multiple). At $\lambda_{2} = 1$, $\mat{A} - \mat{I} =
\begin{psmallmatrix} 4 & 4 \\ 1 & 1 \end{psmallmatrix}$ gives
$\vect{v}_{2} = (1, -1)\transpose$. Take $\mat{S} = \begin{psmallmatrix}
4 & 1 \\ 1 & -1 \end{psmallmatrix}$ and $\mat{\Lambda} = \diag(6, 1)$.
Then $\det \mat{S} = -5$ and $\mat{S}\inv = -\tfrac{1}{5}
\begin{psmallmatrix} -1 & -1 \\ -1 & 4 \end{psmallmatrix} =
\tfrac{1}{5}\begin{psmallmatrix} 1 & 1 \\ 1 & -4 \end{psmallmatrix}$.
Multiplying out $\mat{S} \mat{\Lambda} \mat{S}\inv$ recovers
$\begin{psmallmatrix} 5 & 4 \\ 1 & 2 \end{psmallmatrix}$, confirming
the diagonalisation.

\begin{exercise}
For the symmetric matrix $\mat{A} =
\begin{psmallmatrix} 2 & 1 & 0 \\ 1 & 2 & 1 \\ 0 & 1 & 2
\end{psmallmatrix}$, compute the three eigenvalues and verify that
the eigenvectors are mutually orthogonal. State which mode of a
three-mass linear chain each eigenvector corresponds to.
\end{exercise}

\paragraph{Solution sketch.}
This is the stiffness matrix of section 10.8 with $k = 1$. The
eigenvalues are $\lambda_{1} = 2 - \sqrt{2}$, $\lambda_{2} = 2$, and
$\lambda_{3} = 2 + \sqrt{2}$, found by computing the characteristic
polynomial $-\lambda^{3} + 6\lambda^{2} - 10\lambda + 4$ and
factoring (try $\lambda = 2$ first; the residue is $\lambda^{2} -
4\lambda + 2$, with roots $2 \pm \sqrt{2}$). The unit eigenvectors
are $\vect{v}_{1} = (1, \sqrt{2}, 1)/2$ (in-phase, slow mode),
$\vect{v}_{2} = (1, 0, -1)/\sqrt{2}$ (centre still, medium mode),
and $\vect{v}_{3} = (1, -\sqrt{2}, 1)/2$ (alternating, fast mode).
Pairwise inner products are all zero, confirming orthogonality.
Match the eigenvectors to the modes of \cref{fig:vol02:ch10:vibration-modes}.

\begin{exercise}
For the stochastic matrix $\mat{P} = \begin{psmallmatrix} 0.5 & 0.5
\\ 0.2 & 0.8 \end{psmallmatrix}$, compute the stationary distribution
$\vect{\pi}^{*}$ and the second eigenvalue $\lambda_{2}$. State the
mixing time, defined as $1/(1 - \abs{\lambda_{2}})$.
\end{exercise}

\paragraph{Solution sketch.}
Trace is $1.3$, determinant is $0.4 - 0.1 = 0.3$. Eigenvalues solve
$\lambda^{2} - 1.3\lambda + 0.3 = 0$: $\lambda_{1} = 1$ (stochastic
matrix) and $\lambda_{2} = 0.3$. For the stationary distribution,
solve $\vect{\pi}(\mat{P} - \mat{I}) = 0$ with
$\mat{P} - \mat{I} = \begin{psmallmatrix} -0.5 & 0.5 \\ 0.2 & -0.2
\end{psmallmatrix}$: the first column gives $-0.5\pi_{1} + 0.2\pi_{2}
= 0$, so $\pi_{2}/\pi_{1} = 5/2$. Normalising $\pi_{1} + \pi_{2} = 1$
yields $\vect{\pi}^{*} = (2/7, 5/7) \approx (0.286, 0.714)$. Mixing
time is $1/(1 - 0.3) \approx 1.43$ steps; the chain forgets initial
conditions almost immediately.

\begin{exercise}
Compute the rank-$1$ approximation of $\mat{A} =
\begin{psmallmatrix} 2 & 0 \\ 0 & 1 \\ 0 & 0 \end{psmallmatrix}$ by
truncating its SVD. Verify that the spectral-norm error equals
$\sigma_{2}$ as predicted by Eckart-Young.
\end{exercise}

\paragraph{Solution sketch.}
The matrix is already in diagonal form (up to the third zero row).
Its SVD has $\sigma_{1} = 2$ with $\vect{u}_{1} = (1, 0, 0)\transpose$
and $\vect{v}_{1} = (1, 0)\transpose$; $\sigma_{2} = 1$ with
$\vect{u}_{2} = (0, 1, 0)\transpose$ and $\vect{v}_{2} = (0, 1)
\transpose$. The rank-$1$ truncation is $\mat{A}_{1} = \sigma_{1}
\vect{u}_{1} \vect{v}_{1}\transpose = \begin{psmallmatrix} 2 & 0 \\
0 & 0 \\ 0 & 0 \end{psmallmatrix}$. The error matrix is
$\mat{A} - \mat{A}_{1} = \begin{psmallmatrix} 0 & 0 \\ 0 & 1 \\ 0 & 0
\end{psmallmatrix}$, whose largest singular value is $1 = \sigma_{2}$.
This matches Eckart-Young: $\norm{\mat{A} - \mat{A}_{1}}_{2} =
\sigma_{k+1} = \sigma_{2} = 1$.

\begin{exercise}
For the four data points $(0, 0), (1, 2), (2, 4), (3, 6)$, compute
the centred data matrix, the covariance matrix, the principal
directions, and the explained-variance ratios. State the rank of
the centred data matrix and the geometric meaning of the result.
\end{exercise}

\paragraph{Solution sketch.}
Column means are $\bar{\vect{x}} = (1.5, 3)$. The centred data is
$\mat{X}_{c} = \begin{psmallmatrix} -1.5 & -3 \\ -0.5 & -1 \\ 0.5 &
1 \\ 1.5 & 3 \end{psmallmatrix}$. The covariance matrix is
$\mat{C} = \mat{X}_{c}\transpose \mat{X}_{c}/3 = \begin{psmallmatrix}
5/3 & 10/3 \\ 10/3 & 20/3 \end{psmallmatrix}$. Trace is $25/3$,
determinant is $100/9 - 100/9 = 0$. Eigenvalues are $\lambda_{1} =
25/3$ and $\lambda_{2} = 0$. Principal directions: $\vect{v}_{1} =
(1, 2)/\sqrt{5}$ (along the data line) and $\vect{v}_{2} = (2, -1)/
\sqrt{5}$ (perpendicular). Explained-variance ratios are $1.0$ and
$0$. The centred data matrix has rank $1$: all four centred points
lie exactly on the line through the origin with direction $(1, 2)$.
The geometric meaning is exact one-dimensional structure; PCA
recovers it with zero loss.

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Show that for any square matrix $\mat{A}$, $\tr \mat{A} = \sum_{i}
\lambda_{i}$ and $\det \mat{A} = \prod_{i} \lambda_{i}$, where the
$\lambda_{i}$ are the eigenvalues counted with algebraic multiplicity.
Use the characteristic polynomial.
\end{exercise}

\paragraph{Solution sketch.}
The characteristic polynomial expands as $p(\lambda) = \det(\mat{A} -
\lambda \mat{I}) = (-1)^{n} \prod_{i=1}^{n}(\lambda - \lambda_{i})$.
Expand the product: the coefficient of $\lambda^{n-1}$ on the right
is $(-1)^{n} \cdot (-1) \sum_{i} \lambda_{i} = (-1)^{n+1}
\sum_{i}\lambda_{i}$. On the left, expanding $\det(\mat{A} -
\lambda\mat{I})$ along the diagonal contributes
$(-1)^{n} \lambda^{n} + (-1)^{n-1}(\sum_{i} A_{ii})\lambda^{n-1} +
\cdots$, so the coefficient of $\lambda^{n-1}$ is
$(-1)^{n-1} \tr \mat{A}$. Equating: $\tr \mat{A} = \sum_{i}
\lambda_{i}$. For the determinant, set $\lambda = 0$ on both sides:
$p(0) = \det \mat{A}$ on the left and $(-1)^{n} \prod_{i}(-\lambda_{i})
= \prod_{i} \lambda_{i}$ on the right.

\begin{exercise}
Show that eigenvectors of a symmetric matrix corresponding to
distinct eigenvalues are orthogonal. State the role of the
symmetry assumption in the proof.
\end{exercise}

\paragraph{Solution sketch.}
Let $\mat{A} = \mat{A}\transpose$, with $\mat{A} \vect{v}_{1} =
\lambda_{1} \vect{v}_{1}$ and $\mat{A} \vect{v}_{2} = \lambda_{2}
\vect{v}_{2}$, $\lambda_{1} \neq \lambda_{2}$. Then
$\lambda_{1} \vect{v}_{2}\transpose \vect{v}_{1} = \vect{v}_{2}
\transpose \mat{A} \vect{v}_{1} = (\mat{A}\transpose \vect{v}_{2})
\transpose \vect{v}_{1} = (\mat{A} \vect{v}_{2})\transpose
\vect{v}_{1} = \lambda_{2} \vect{v}_{2}\transpose \vect{v}_{1}$.
So $(\lambda_{1} - \lambda_{2}) \vect{v}_{2}\transpose \vect{v}_{1}
= 0$, which forces $\vect{v}_{2}\transpose \vect{v}_{1} = 0$. The
symmetry assumption $\mat{A}\transpose = \mat{A}$ is essential at
the step where we move $\mat{A}$ across the inner product; without
symmetry, $\mat{A}\transpose \vect{v}_{2}$ would not equal $\mat{A}
\vect{v}_{2}$ and the cancellation fails.

\begin{exercise}
Derive the Eckart-Young theorem in the Frobenius norm: starting from
the SVD $\mat{A} = \mat{U} \mat{\Sigma} \mat{V}\transpose$, show
that the rank-$k$ truncation $\mat{A}_{k}$ minimises $\norm{\mat{A}
- \mat{B}}_{F}^{2}$ over rank-at-most-$k$ matrices $\mat{B}$, and
that the optimal value is $\sum_{i=k+1}^{p} \sigma_{i}^{2}$.
\end{exercise}

\paragraph{Solution sketch.}
The Frobenius norm is unitarily invariant: $\norm{\mat{U}\mat{M}
\mat{V}\transpose}_{F} = \norm{\mat{M}}_{F}$ for any orthogonal
$\mat{U}, \mat{V}$. Substitute: $\norm{\mat{A} - \mat{B}}_{F}^{2} =
\norm{\mat{\Sigma} - \mat{U}\transpose \mat{B} \mat{V}}_{F}^{2} =
\norm{\mat{\Sigma} - \mat{C}}_{F}^{2}$, with $\mat{C} = \mat{U}
\transpose \mat{B} \mat{V}$ of rank at most $k$. Now $\norm{\mat{\Sigma}
- \mat{C}}_{F}^{2} = \sum_{i,j}(\Sigma_{ij} - C_{ij})^{2}$. Since
$\mat{\Sigma}$ is diagonal, $\sum_{i \neq j} C_{ij}^{2} +
\sum_{i}(\sigma_{i} - C_{ii})^{2}$. To minimise, set all
off-diagonals of $\mat{C}$ to zero. The problem reduces to choosing
a vector $(c_{1}, \ldots, c_{p})$ with at most $k$ nonzero entries
minimising $\sum_{i}(\sigma_{i} - c_{i})^{2}$. The optimum is
$c_{i} = \sigma_{i}$ for the $k$ largest $\sigma_{i}$ (i.e.\
$i = 1, \ldots, k$) and $c_{i} = 0$ otherwise. The optimal value is
$\sum_{i=k+1}^{p} \sigma_{i}^{2}$, achieved by $\mat{C} = \diag(\sigma_{1},
\ldots, \sigma_{k}, 0, \ldots, 0)$, equivalently
$\mat{B} = \mat{A}_{k}$.

\begin{exercise}
Show that a stochastic matrix $\mat{P}$ has $\lambda = 1$ as an
eigenvalue, with the all-ones vector as a right eigenvector. Using
the Gerschgorin circle theorem, show that every eigenvalue
satisfies $\abs{\lambda} \leq 1$.
\end{exercise}

\paragraph{Solution sketch.}
For the right eigenvector: $(\mat{P} \vect{1})_{i} = \sum_{j} P_{ij}
\cdot 1 = \sum_{j} P_{ij} = 1 = 1 \cdot \vect{1}_{i}$, where the
penultimate equality uses the row-sum-one property of stochastic
matrices. So $\mat{P} \vect{1} = \vect{1}$ and $\lambda = 1$ is an
eigenvalue. For the bound: by Gerschgorin, every eigenvalue lies in
the union of discs centred at $P_{ii}$ with radius $\sum_{j \neq i}
\abs{P_{ij}} = 1 - P_{ii}$ (using non-negativity and the row sum).
Each disc is contained in the closed unit disc: the centre $P_{ii}$
is real and in $[0, 1]$, and the radius is $1 - P_{ii}$, so the
farthest point in the disc from the origin is $P_{ii} + (1 - P_{ii})
= 1$. Hence $\abs{\lambda} \leq 1$ for every eigenvalue.

\begin{exercise}
Show that the SVD of a real matrix $\mat{A}$ is related to the
eigendecomposition of $\mat{A}\transpose \mat{A}$ by:
the singular values are the square roots of the eigenvalues of
$\mat{A}\transpose \mat{A}$, and the right singular vectors are the
corresponding eigenvectors. State why this relation does not, by
itself, give a numerically stable SVD algorithm.
\end{exercise}

\paragraph{Solution sketch.}
From $\mat{A} = \mat{U} \mat{\Sigma} \mat{V}\transpose$ compute
$\mat{A}\transpose \mat{A} = \mat{V} \mat{\Sigma}\transpose
\mat{U}\transpose \mat{U} \mat{\Sigma} \mat{V}\transpose = \mat{V}
(\mat{\Sigma}\transpose \mat{\Sigma}) \mat{V}\transpose$. The matrix
$\mat{\Sigma}\transpose \mat{\Sigma}$ is $n \times n$ diagonal with
entries $\sigma_{i}^{2}$. This is the eigendecomposition of
$\mat{A}\transpose \mat{A}$ with eigenvalues $\sigma_{i}^{2}$ and
eigenvectors the columns of $\mat{V}$. The numerical concern: forming
$\mat{A}\transpose \mat{A}$ squares the condition number,
$\kappa(\mat{A}\transpose \mat{A}) = \kappa(\mat{A})^{2}$. A matrix
with $\kappa(\mat{A}) = 10^{6}$ becomes $\kappa(\mat{A}\transpose
\mat{A}) = 10^{12}$, well into the regime where double-precision
arithmetic loses meaningful digits. Production SVD (Golub-Reinsch,
1970, \cite{paper:v2ch10-golub-reinsch1970}) works on $\mat{A}$
directly by bidiagonal reduction and avoids the squaring.

\begin{exercise}
Show that for a diagonalisable $\mat{A} = \mat{S} \mat{\Lambda}
\mat{S}\inv$, the matrix exponential is $e^{\mat{A} t} = \mat{S}
e^{\mat{\Lambda} t} \mat{S}\inv$, where $e^{\mat{\Lambda} t}$ is
the diagonal matrix with entries $e^{\lambda_{i} t}$. Verify by
substitution into the series definition $e^{\mat{A} t} = \sum_{k}
(\mat{A} t)^{k}/k!$.
\end{exercise}

\paragraph{Solution sketch.}
Compute $\mat{A}^{k} = (\mat{S} \mat{\Lambda} \mat{S}\inv)^{k} =
\mat{S} \mat{\Lambda} \mat{S}\inv \cdot \mat{S} \mat{\Lambda}
\mat{S}\inv \cdots \mat{S} \mat{\Lambda} \mat{S}\inv = \mat{S}
\mat{\Lambda}^{k} \mat{S}\inv$ (the inner $\mat{S}\inv \mat{S}$
factors cancel). Substitute into the series: $e^{\mat{A} t} =
\sum_{k=0}^{\infty} (\mat{A} t)^{k}/k! = \sum_{k} \mat{S} (\mat{\Lambda}
t)^{k} \mat{S}\inv / k! = \mat{S} \left(\sum_{k} (\mat{\Lambda} t)^{k}
/ k!\right) \mat{S}\inv = \mat{S} e^{\mat{\Lambda} t} \mat{S}\inv$.
The last sum is diagonal because $\mat{\Lambda}$ is diagonal, with
$(e^{\mat{\Lambda} t})_{ii} = \sum_{k} (\lambda_{i} t)^{k}/k! =
e^{\lambda_{i} t}$.

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the number of power-iteration steps required to compute
the dominant eigenvalue of a matrix with eigenvalue ratio
$\abs{\lambda_{2}/\lambda_{1}} = 0.9$, to three decimal digits of
accuracy. State how the estimate changes if the ratio is $0.5$ or
$0.99$.
\end{exercise}

\paragraph{Solution sketch.}
Power iteration's error after $t$ steps is approximately
$r^{t}$ with $r = \abs{\lambda_{2}/\lambda_{1}}$. Three decimal
digits of accuracy means $r^{t} \leq 10^{-3}$, so $t \geq -3 \log
10 / \log r$. For $r = 0.9$: $t \geq 3 \cdot 2.303 / 0.105 \approx
66$ steps. For $r = 0.5$: $t \geq 3 \cdot 2.303 / 0.693 \approx 10$
steps. For $r = 0.99$: $t \geq 3 \cdot 2.303 / 0.01005 \approx 687$
steps. The penalty for a closing spectral gap is brutal: a tenfold
increase in $1 - r$ buys a sevenfold reduction in iterations. Below
$r = 0.99$, the engineer should use a more sophisticated method
(Lanczos for symmetric, Arnoldi for general).

\begin{exercise}
Estimate the rank at which a $1024 \times 1024$ grayscale photograph
is visually indistinguishable from the original after truncated-SVD
reconstruction, given that natural-image singular values typically
decay roughly as $\sigma_{k} \propto k^{-1}$ over the first few
hundred. State the storage ratio achieved at this rank.
\end{exercise}

\paragraph{Solution sketch.}
Visual indistinguishability typically requires
$\sigma_{k}/\sigma_{1} \leq 1\%$, i.e.\ $k \geq 100$ for a $1/k$
decay. Be conservative and take $k = 200$ to leave headroom for the
slower tail. The truncated storage is $k(m + 1 + n) \approx 200
\cdot 2049 = 4.1 \times 10^{5}$ entries; the full storage is
$1024^{2} \approx 1.05 \times 10^{6}$. The storage ratio is
$4.1/10.5 \approx 0.39$, a $2.6 \times$ reduction. The compression
is modest because the Eckart-Young bound is sharp and natural
images decay slowly. For images of text or with sharp edges the
decay is even slower and the effective compression drops further.
For low-rank-plus-noise data (a few-component PCA-like signal),
$k \approx 10$ may give a $50 \times$ compression.
\Cref{fig:vol02:ch10:singular-value-decay} makes the data-class
dependence visible.

\begin{exercise}
Estimate the second-largest eigenvalue magnitude of the PageRank
matrix at teleportation parameter $\alpha = 0.85$, given the
spectral-gap bound $\abs{\lambda_{2}} \leq \alpha$. State the number
of power-iteration steps required for $10^{-6}$ accuracy and
compare to the number for $\alpha = 0.99$.
\end{exercise}

\paragraph{Solution sketch.}
The spectral-gap bound gives $\abs{\lambda_{2}} \leq 0.85$. Power
iteration converges at rate $\abs{\lambda_{2}/\lambda_{1}} \leq 0.85
/1 = 0.85$. For $10^{-6}$ accuracy: $0.85^{t} \leq 10^{-6}$ gives
$t \geq 6 \log 10 / (-\log 0.85) \approx 13.82/0.1625 \approx 85$
steps. For $\alpha = 0.99$: $0.99^{t} \leq 10^{-6}$ gives $t \geq
13.82/0.01005 \approx 1376$ steps, a $16 \times$ increase. The
default $\alpha = 0.85$ is calibrated to give the rate-versus-
faithfulness compromise that web-scale ranking can afford. A larger
$\alpha$ would rank more faithfully to the link structure at large
cost in compute.

\begin{exercise}
Estimate the natural frequencies of a five-mass equal-spring
linear chain (six equal springs of stiffness $k$, five equal masses
$m$, walls at both ends). State the highest and lowest frequencies
in units of $\sqrt{k/m}$, by reading them off the limiting cases
(all in phase, alternating sign), without computing the full
spectrum.
\end{exercise}

\paragraph{Solution sketch.}
The stiffness matrix is the $5 \times 5$ analogue of the three-mass
case, with $2$ on the diagonal and $-1$ on the sub- and
super-diagonals (times $k$). The eigenvalues are known in closed
form: $\lambda_{j} = 2(1 - \cos(j\pi/6))$ for $j = 1, \ldots, 5$.
The lowest is $\lambda_{1} = 2(1 - \cos 30^{\circ}) = 2(1 -
\sqrt{3}/2) \approx 0.268$, giving $\omega_{1} \approx 0.518
\sqrt{k/m}$. The highest is $\lambda_{5} = 2(1 - \cos 150^{\circ}) =
2(1 + \sqrt{3}/2) \approx 3.732$, giving $\omega_{5} \approx 1.932
\sqrt{k/m}$. The ratio $\omega_{5}/\omega_{1} \approx 3.7$ is the
chain's frequency spread, much wider than the three-mass system's
$\sqrt{3+\sqrt{2}}/\sqrt{2-\sqrt{2}} \approx 2.4$. As the chain
grows, $\omega_{1} \to 0$ (lowest mode wavelength approaches the
full chain length) and $\omega_{\max} \to 2\sqrt{k/m}$ (alternating
neighbours, the discrete-lattice cutoff).

\begin{exercise}
Estimate the proportion of the variance captured by the first
principal component of a dataset whose feature correlation
coefficient is $0.9$ across all features. State the proportion as
a function of the number of features $n$ and verify the limiting
behaviour as $n \to \infty$.
\end{exercise}

\paragraph{Solution sketch.}
A correlation matrix $\mat{R}$ with $1$ on the diagonal and $\rho$
everywhere off-diagonal has eigenvalues $\lambda_{1} = 1 + (n-1)\rho$
(eigenvector all-ones) and $\lambda_{2} = \cdots = \lambda_{n} =
1 - \rho$ (eigenvectors any orthogonal complement of the all-ones
direction). The trace is $n$, so the explained-variance ratio of
the first PC is $\lambda_{1}/n = (1 + (n-1)\rho)/n$. With $\rho =
0.9$ and $n = 10$: $(1 + 9 \cdot 0.9)/10 = 0.91$. With $n = 100$:
$(1 + 99 \cdot 0.9)/100 = 0.901$. As $n \to \infty$: $(1 + (n-1)
\rho)/n \to \rho = 0.9$. The proportion converges to the
correlation coefficient itself, a clean limit that justifies the
practitioner's rule that highly correlated features compress almost
entirely into one PC.

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement power iteration on a $50 \times 50$ random symmetric
matrix. Plot the convergence of the Rayleigh quotient to the
dominant eigenvalue on a log-y axis. Verify the geometric
convergence rate $\abs{\lambda_{2}/\lambda_{1}}^{t}$ predicted by
section 10.6.
\end{exercise}

\paragraph{Solution sketch.}
The reference implementation \texttt{code/power\_iteration.py}
covers this exercise directly. Draw a random symmetric matrix with
Gaussian entries; compute the reference eigenvalues with a library
eigensolver; track the Rayleigh quotient $\vect{x}_{t}\transpose
\mat{A} \vect{x}_{t}$ across iterations and the error
$\abs{\lambda_{1}^{(t)} - \lambda_{1}^{\text{true}}}$. Plot the
error against $t$ on a log-y axis. The error decays as a straight
line with slope $\log_{10}\abs{\lambda_{2}/\lambda_{1}}$, matching
the principle in section 10.6. For a $50 \times 50$ Gaussian-symmetric
matrix the ratio is typically near $0.9$, so a few hundred iterations
buy three to four digits.

\begin{exercise}
Implement deflation: after computing the dominant eigenpair
$(\lambda_{1}, \vect{v}_{1})$ by power iteration, subtract
$\lambda_{1} \vect{v}_{1} \vect{v}_{1}\transpose$ from $\mat{A}$
(for symmetric $\mat{A}$) and run power iteration on the deflated
matrix to find the next eigenpair. Verify on a $20 \times 20$
symmetric random matrix and compare to a library eigensolver.
\end{exercise}

\paragraph{Solution sketch.}
The \texttt{--deflate} flag in \texttt{code/power\_iteration.py}
runs the deflation. After computing $\lambda_{1}, \vect{v}_{1}$,
form $\mat{A}_{1} = \mat{A} - \lambda_{1} \vect{v}_{1}
\vect{v}_{1}\transpose$. By construction, $\mat{A}_{1}$ has the
same eigenvectors as $\mat{A}$ but with eigenvalue $0$ at
$\vect{v}_{1}$ and unchanged eigenvalues elsewhere. Power iteration
on $\mat{A}_{1}$ converges to the eigenvector at $\lambda_{2}$
(now the dominant). Verify by comparing to the library
eigendecomposition. Caveat: a small residual component of
$\vect{v}_{1}$ in the deflated iterate can re-grow if $\lambda_{1}$
is large; use re-orthogonalisation against $\vect{v}_{1}$ at each
step to stabilise.

\begin{exercise}
Implement the truncated-SVD compression of a grayscale image of
your choice (any $\geq 256 \times 256$ photograph). Display the
reconstruction at $k = 10, 50, 100, 200$ and report the
spectral-norm error at each rank against the Eckart-Young
prediction $\sigma_{k+1}$.
\end{exercise}

\paragraph{Solution sketch.}
The reference implementation \texttt{code/svd\_image\_compression.py}
runs this exercise on any grayscale or colour image, reporting both
the spectral-norm and Frobenius-norm errors at each rank against
the Eckart-Young predictions. At $k = 10$ the reconstruction is
blurry but recognisable for portraits; at $k = 50$ it is acceptable
for casual viewing; at $k = 100$ to $200$ it is visually
indistinguishable from the original for most natural images. The
spectral-norm error matches $\sigma_{k+1}$ to all reported digits
(this is an identity, not an approximation, when the SVD is computed
to working precision). Report the storage ratio at each rank; the
break-even (where truncated storage equals original storage) is
near $k \approx mn/(m+n+1)$.

\begin{exercise}
Simulate a Markov chain on a 5-state cycle with one ``leaky''
transition (state $i$ goes to state $i+1$ with probability $0.9$,
to state $1$ with probability $0.1$). Compute the stationary
distribution by both eigendecomposition and direct power iteration
on the row vector. Verify they agree to four decimal digits and
report the mixing time.
\end{exercise}

\paragraph{Solution sketch.}
The reference implementation \texttt{code/markov\_stationary.py}
runs this exercise directly. The chain is irreducible (the leak to
state $1$ guarantees return) and aperiodic (state $1$ has a self-
return probability of $0.1 \cdot 0$ at one step but the leak
introduces variable return paths whose gcd is $1$). Both methods
return the same stationary distribution to machine precision; the
second-largest eigenvalue magnitude $\abs{\lambda_{2}}$ governs
the mixing time, approximately $1/(1 - \abs{\lambda_{2}})$. For the
specified chain the leak is $0.1$, so the mixing time is on the
order of $10$ steps. Reducing the leak to $0.01$ would push the
mixing time toward $100$ and bring power iteration close to the
useful-iteration boundary.

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a procedure for choosing the number of principal components
$k$ in a PCA pipeline, given a dataset and a downstream task. The
procedure should consider the explained-variance ratio, the
scree-plot elbow, and the downstream task's sensitivity to
truncation noise. State the criteria used and justify each.
\end{exercise}

\paragraph{Discussion.}
Three criteria, applied jointly. First, the explained-variance
threshold: pick $k$ such that the cumulative explained-variance
ratio reaches a chosen level (typical engineering defaults are
$90\%$ for visualisation, $95\%$ for storage, $99\%$ for downstream
predictive modelling). Second, the scree-plot elbow: plot
$\sigma_{i}^{2}$ versus $i$ and pick $k$ at the visual elbow where
the curve transitions from steep decay to slow tail. Third, the
downstream task's sensitivity: run the downstream task (classifier,
regressor, controller) at several candidate $k$ values and pick the
smallest $k$ that does not degrade the task metric beyond a chosen
tolerance. Justify each criterion: the first is a global energy
argument; the second is a structural argument about the data's
intrinsic rank; the third is an empirical guarantee that the
truncation does not lose downstream information. In tension, the
third dominates: the goal is the task, not the variance ratio.

\begin{exercise}
Design a power-iteration variant for the case where the dominant
eigenvalue is negative or complex. State the modification (shift
to make all eigenvalues positive, or work with $\mat{A}\transpose
\mat{A}$ instead of $\mat{A}$) and the trade-offs in convergence
rate and computational cost.
\end{exercise}

\paragraph{Discussion.}
Two principled variants. (a) Shifted iteration: pick $\sigma$
larger than the spectral radius of $-\mat{A}$ and iterate
$(\mat{A} + \sigma \mat{I})$. All eigenvalues of the shifted matrix
are positive and the dominant one is now reliably the largest in
magnitude. The cost is one extra addition per iteration; the
convergence rate is $\abs{(\lambda_{2} + \sigma)/(\lambda_{1} +
\sigma)}$, which approaches $1$ as $\sigma$ grows large, so $\sigma$
should be the smallest value that makes all eigenvalues positive.
(b) Iterate on $\mat{A}\transpose \mat{A}$: the eigenvalues are
squared and all positive, the eigenvectors are the right singular
vectors of $\mat{A}$. The cost is one extra matrix-vector multiply
per iteration (two instead of one) and squaring of the spectral
gap; the convergence rate is $(\lambda_{2}/\lambda_{1})^{2}$, which
is faster than the original whenever $\abs{\lambda_{2}/\lambda_{1}}
< 1$. For complex dominant eigenvalues the second variant is
preferred because it recovers a real iterate; the first variant
still leaves a complex dominant pair.

\begin{exercise}
Design a numerical pipeline for computing the matrix exponential
$e^{\mat{A} t}$ for a $1000 \times 1000$ non-symmetric $\mat{A}$
that may be near-defective. State which of the following you would
use and why: scaling-and-squaring with Pad\'e approximation,
diagonalisation, Schur-based function evaluation, or Krylov
subspace projection.
\end{exercise}

\paragraph{Discussion.}
Use scaling-and-squaring with Pad\'e approximation. The reasoning:
diagonalisation amplifies error by $\kappa(\mat{S})$ where
$\mat{S}$ is the eigenvector matrix, and a near-defective $\mat{A}$
gives $\kappa(\mat{S})$ much larger than $\kappa(\mat{A})$. Schur-
based function evaluation (Parlett recurrence) works for general
matrices but is more expensive at $\oom{n^{3}}$ per matrix-function
call and has its own conditioning issues at close eigenvalues.
Krylov subspace projection is the right tool for $\mat{A}$ very
large and sparse with the action $e^{\mat{A} t} \vect{b}$ on a
single vector $\vect{b}$ needed; at $n = 1000$ the dense
scaling-and-squaring is competitive. Scaling-and-squaring with Pad\'e
approximation is the Higham-recommended default
\cite[ch.~10]{text:v2ch10-higham-functions2008}; it is what
\texttt{expm} in NumPy and MATLAB call by default, with conditioning
robust to near-defective matrices.

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A colleague reports that a $20 \times 20$ matrix has condition
number $10^{12}$ but that its eigenvalue decomposition produced a
clean eigenvalue spectrum with no obvious near-degeneracy. Identify
where the trouble is hiding and propose two diagnostics that would
locate it.
\end{exercise}

\paragraph{Solution sketch.}
Condition number measures singular-value ratio, not
eigenvalue-ratio. The matrix can have well-separated eigenvalues
but a poorly conditioned eigenvector matrix $\mat{S}$: the trouble
is non-normality, $\mat{A}\transpose \mat{A} \neq \mat{A}\mat{A}
\transpose$. Two diagnostics: (i) compute $\norm{\mat{A}\transpose
\mat{A} - \mat{A} \mat{A}\transpose}_{F}/\norm{\mat{A}}_{F}^{2}$;
if the ratio exceeds machine precision, the matrix is non-normal
and the eigenvalues under-describe its behaviour. (ii) compute the
condition number of $\mat{S}$ itself,
$\kappa(\mat{S})$; if it is large (say $\geq 10^{6}$), the
eigenvectors are nearly parallel and the diagonalisation
$\mat{A} = \mat{S} \mat{\Lambda} \mat{S}\inv$ amplifies error by
$\kappa(\mat{S})$. Use the Schur decomposition for downstream
computations instead.

\begin{exercise}
A PageRank computation on a web graph of $10^{6}$ nodes appears to
diverge: the iterate norm grows by a small factor each step.
Diagnose the most likely cause (dangling nodes, disconnected
components, off-by-one in the teleportation correction, or
floating-point overflow in a large entry) and propose the test
that distinguishes them.
\end{exercise}

\paragraph{Solution sketch.}
Most likely cause: dangling-node correction not applied. Without
the correction $\mat{H}$ has rows summing to less than one, so
$\mat{H}$ is sub-stochastic and the iterate loses mass each step.
Diagnostic: compute the row sums of the transition matrix being
iterated; if any row sums to anything other than $1$ (to machine
precision), the dangling correction is missing or the
teleportation has an off-by-one. Floating-point overflow in a large
entry is unlikely at the scale described because the iterate is
normalised at each step in any sane implementation. Disconnected
components would not cause divergence but would produce multiple
stationary distributions or block-diagonal structure; check by
running a strongly-connected-components algorithm on the graph and
verifying it returns a single component (with the teleportation
correction applied).

\begin{exercise}
A modal-analysis software returns two natural frequencies of a
mechanical structure within $0.01\%$ of each other, although the
structure is asymmetric. Diagnose whether this represents a
genuine near-degenerate mode pair, a numerical artefact, or a
modelling error in the input file. Propose three checks to
distinguish.
\end{exercise}

\paragraph{Solution sketch.}
Three checks. (i) Inspect the mode shapes. A genuine near-
degenerate pair has two distinct mode shapes that are orthogonal;
a numerical artefact produces two near-parallel shapes (the solver
has split a single mode in two). Compute
$\vect{v}_{i}\transpose \vect{v}_{j}/(\norm{\vect{v}_{i}}
\norm{\vect{v}_{j}})$; if it is near zero the modes are genuine. (ii)
Perturb the stiffness matrix by a small symmetry-breaking term and
re-solve; the two frequencies should move continuously. If they
collapse to one mode the original split was a numerical artefact.
(iii) Inspect the input file for accidental symmetries (mirror
geometry, equal material properties in both directions) that the
``asymmetric'' label did not capture; modelling error often hides
in repeated material constants. The combination of mode-shape
orthogonality and perturbation continuity is the conclusive test.

\begin{exercise}
A PCA on a dataset reports the first principal component captures
$99.9\%$ of variance, but the second through fifth principal
directions appear to be permutations of the same vector with sign
flips. Identify the most likely data-quality problem and propose
the diagnostic check.
\end{exercise}

\paragraph{Solution sketch.}
Most likely cause: one feature has variance many orders of
magnitude larger than the others, dominating the covariance matrix
and producing one strong PC with the remaining components being
numerical noise. (Equivalently: PCA was applied to unscaled
heterogeneous-unit features, e.g.\ mixing kilograms and millimetres
in one matrix.) Diagnostic: compute the per-feature variance
$\Var(\mat{X}_{c}[:, j])$; if one variance dominates the others by
$\geq 10^{4}\times$, the explained-variance ratio is a unit-
sensitivity artefact rather than a structural finding. The standard
remedy is to standardise: divide each feature by its standard
deviation before centring, so the covariance becomes the correlation
matrix and the PCs reflect linear structure rather than unit choice.

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Argue, in 200 to 300 words, why naive diagonalisation is the wrong
tool for computing the matrix exponential of a near-defective
matrix, even when a library eigensolver returns ostensibly clean
eigenvalues. Identify the specific error-amplification mechanism
and the production-grade alternative.
\end{exercise}

\paragraph{Solution sketch.}
A near-defective matrix has nearly parallel eigenvectors. The
eigenvector matrix $\mat{S}$ then has $\kappa_{2}(\mat{S}) \gg 1$,
even though the eigenvalues themselves are well separated and
look clean. The matrix exponential by diagonalisation evaluates
$e^{\mat{A} t} = \mat{S} e^{\mat{\Lambda} t} \mat{S}\inv$. The
forward-error analysis of this computation has a multiplicative
factor of $\kappa(\mat{S})$ on the floating-point error of each
component. A matrix with $\kappa(\mat{S}) = 10^{10}$ amplifies
single-precision noise ($\sim 10^{-7}$) to magnitude $10^{3}$,
swamping the true exponential. The eigenvalues' cleanness gives no
warning; the error lives in the change-of-basis, not in the
spectrum. The production-grade alternative is scaling-and-squaring
with Pad\'e approximation \cite[ch.~10]{text:v2ch10-higham-functions2008},
which works on $\mat{A}$ directly via the identity $e^{\mat{A}} =
(e^{\mat{A}/2^{s}})^{2^{s}}$ and the Pad\'e approximation to
$e^{\mat{B}}$ for $\mat{B}$ with small norm. The algorithm has
backward stability comparable to working precision regardless of the
defectiveness of $\mat{A}$. It is what \texttt{scipy.linalg.expm}
and \texttt{expm} in MATLAB call by default. The engineer should
treat the existence of $\mat{S}\inv$ as a precondition to be
checked, not assumed.

\begin{exercise}
A control system's closed-loop transfer matrix has eigenvalues that
the engineer computes as $-0.01 \pm 0.5 i$, classifying the system
as stable. The system, when deployed, exhibits growing oscillation
and fails. Argue, in 200 to 300 words, what numerical or modelling
failure could have produced a stable-looking spectrum from an
unstable system, and propose three checks the engineer should have
performed before deployment.
\end{exercise}

\paragraph{Solution sketch.}
Three plausible causes. (a) Non-normality and large transient
growth: even with eigenvalues in the left half-plane, a strongly
non-normal closed-loop matrix can produce transient growth of
$\norm{e^{\mat{A} t} \vect{x}_{0}}$ by orders of magnitude before
decay sets in. If the plant's saturation or actuator limits are
exceeded during the transient, the linear analysis no longer
predicts behaviour and a nonlinear instability ensues. (b) Modelling
mismatch: the closed-loop transfer matrix used for eigenvalue
analysis was the nominal model, but the deployed plant differs
(parameter drift, sensor latency, unmodelled mechanical mode). The
true eigenvalues are then different and may sit in the right half-
plane. (c) Discretisation error: a continuous-time stability
analysis was applied to a system that is digitally controlled at a
sampling rate where the discrete-time eigenvalues sit outside the
unit disc even when the continuous eigenvalues are in the left half-
plane.

Three pre-deployment checks. First, compute the pseudospectrum of
the closed-loop matrix and verify the $\epsilon$-pseudospectrum for
realistic $\epsilon$ stays in the left half-plane. Second, run a
robustness analysis: perturb the plant parameters by their stated
tolerances and recompute the spectrum; report worst-case eigenvalue
locations. Third, simulate the closed loop in discrete time at the
intended sampling rate and verify the discrete-time eigenvalues sit
inside the unit disc with adequate margin.

\begin{exercise}
A finite-element analysis of a three-dimensional truss uses
diagonalisation of the stiffness matrix to compute modal responses.
The analysis returns six natural frequencies at exactly zero
(within machine precision). Argue, in 200 to 300 words, what these
zeros mean physically, why they are not a numerical bug, and how
the engineer should treat them in the modal superposition.
\end{exercise}

\paragraph{Solution sketch.}
A free-floating three-dimensional rigid body has six rigid-body
modes: three translations and three rotations. The stiffness matrix
in the global coordinates is singular along these modes because
rigid-body motion produces zero strain and thus zero restoring
force. Six zero eigenvalues are exactly the right answer; the
truss is not yet attached to anything.

The zeros are not a numerical bug because the algorithm has
returned the exact mathematical answer: $\det \mat{K} = 0$, with
nullity six (rigid translations and rotations form a
six-dimensional null space). A floating-point computation will
return them as values on the order of $\norm{\mat{K}}_{2} \cdot
\epsilon_{\text{machine}}$, which is the algorithm's correct
expression of ``zero.''

In modal superposition the rigid-body modes carry the truss as a
whole through space and contribute no oscillation. The engineer
should treat them in one of two ways. (a) Apply boundary conditions
that pin enough degrees of freedom to remove the rigid-body modes
(typically three constrained translations plus three rotations).
$\mat{K}$ becomes positive definite and the modal analysis returns
only oscillatory modes. (b) Retain the rigid-body modes and project
the loading onto the elastic modes only, computing the truss's
elastic response in the centre-of-mass frame; the rigid-body
response is then computed separately by Newton-Euler dynamics. Both
are standard; the first is more common in structural codes, the
second in multibody dynamics.

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A vendor reports a PCA-based dimensionality-reduction pipeline that
retains $99\%$ of variance with $k = 5$ components on a dataset
with $n = 100$ features. Argue, in one paragraph, whether the
report is meaningful as stated and what the engineer should ask
the vendor before adopting the pipeline for downstream design.
\end{exercise}

\paragraph{Discussion.}
The claim is not meaningful as stated. Variance retention depends
on the units of the features. A dataset with one
high-variance feature in pascals and 99 low-variance features in
millimetres will produce a one-PC pipeline that captures $99\%$ of
variance trivially, with no genuine compression of the underlying
structure. The engineer should ask: (i) Were the features
standardised (zero mean, unit variance) before PCA? If not, the
$99\%$ figure is a unit artefact. (ii) What is the explained-variance
ratio at $k = 1, 2, 3, 4$ separately, not just the cumulative at
$k = 5$? A sharp drop after $k = 1$ means one dominant feature;
a smooth decay means $k = 5$ is doing real compression. (iii) Was the
PCA fit on the training set only and applied to test data, or were
both pooled? Information leakage inflates apparent retention. (iv) Does
the downstream task degrade if $k$ is reduced further, and by how
much? The right metric is task degradation, not variance retention.

\begin{exercise}
A research paper reports that a large language model's
embedding matrix has effective rank around $300$, based on the
truncated-SVD energy threshold at $99\%$. Argue, in 200 to 300
words, what claims this rank estimate supports, what claims it does
not support, and what additional evidence would be needed to make
the claim load-bearing for downstream design.
\end{exercise}

\paragraph{Discussion.}
What the claim supports. The embedding matrix is approximately
captured to $99\%$ Frobenius energy by a rank-$300$ truncated SVD;
storing the rank-$300$ approximation costs about $300 (m + n)$
floats instead of $mn$; downstream operations whose accuracy
depends on Frobenius-norm proximity (some retrieval applications,
some unsupervised clustering schemes) can use the rank-$300$
approximation with bounded loss.

What the claim does not support. The embedding matrix is not
literally rank $300$; the tail $1\%$ of energy may carry essential
information for tasks that depend on small-magnitude directions.
Tasks sensitive to spectral norm (some optimisation routines, some
robustness analyses) are not bounded by the Frobenius-norm energy
threshold and may degrade much more than $1\%$. The figure does not
say anything about the embedding's expressiveness, semantic
content, or capacity for downstream fine-tuning.

Additional evidence needed. (i) Task-degradation curves on a panel
of downstream evaluations, plotting accuracy as a function of $k$.
The relevant $k$ is the smallest at which the panel does not
degrade beyond a chosen tolerance, not the one with $99\%$ energy.
(ii) Spectral-norm error $\sigma_{k+1}$ alongside Frobenius energy,
to detect spectral-norm-sensitive failures. (iii) Per-class or
per-token error analysis on the truncated embedding; rare tokens
often live in low-energy directions and are first to be lost. The
energy-threshold heuristic is a starting point, not a deployment
criterion.

\begin{exercise}
An engineer is offered two implementations of an eigensolver: a
QR-algorithm-based dense solver, and a Lanczos-based sparse solver.
The application is a $10^{5} \times 10^{5}$ symmetric matrix
with $10^{6}$ nonzeros. Argue, in 200 to 300 words, which
implementation the engineer should choose, and state the matrix
size and density at which the choice would flip.
\end{exercise}

\paragraph{Discussion.}
Choose Lanczos. The matrix has $10^{6}$ nonzeros out of $10^{10}$
possible entries, a density of $10^{-4}$: it is sparse. The dense
QR algorithm costs $\oom{n^{3}} = 10^{15}$ flops and $\oom{n^{2}} =
10^{10}$ floats of storage to materialise the dense matrix, which
alone exceeds typical RAM. Lanczos uses only matrix-vector products
on the sparse structure, costing $\oom{\text{nnz}}$ per iteration
and converging in a small multiple of the desired number of
eigenvalues. For computing the lowest few or highest few
eigenvalues (the usual interest), Lanczos finishes in seconds; the
dense QR cannot start.

The choice flips when the matrix becomes dense enough that storing
it densely is comparable to its sparse representation, and when
asking for many eigenvalues (most or all). The crossover density is
roughly $\rho \sim 1/n$: at $n = 10^{5}$ this is $10^{-5}$, so
densities above $10^{-5}$ favour the dense solver if all
eigenvalues are needed. The crossover in matrix size is around
$n \sim 10^{3}$ for dense matrices on commodity hardware: a $10^{3}
\times 10^{3}$ dense eigendecomposition takes seconds and dense
solvers are simpler; below that, the algorithm choice is dominated
by implementation overhead and the dense path is usually preferred.
For very large sparse matrices ($n \geq 10^{5}$) Lanczos or one of
its restarted variants is the only viable option.

\begin{exercise}
A reader of this chapter argues that, since the SVD covers every
matrix and the Eckart-Young theorem makes truncation optimal in two
norms, the SVD should be the default tool for every linear-algebra
problem. Argue, in one paragraph, why this is wrong, identifying at
least two situations where a different decomposition (LU, QR, Schur,
Cholesky) is the right default and stating why the SVD is not.
\end{exercise}

\paragraph{Discussion.}
SVD is universal but expensive: $\oom{n^{3}}$ flops with a constant
several times that of competing decompositions. For solving linear
systems $\mat{A} \vect{x} = \vect{b}$ on a square invertible matrix,
LU factorisation with partial pivoting is the default: it costs
$\oom{n^{3}}/3$ flops (less than half of SVD), is backward stable
in practice, and produces a sparse factorisation when $\mat{A}$ is
sparse with good ordering. For overdetermined least squares with
$\mat{A}$ tall and full column rank, QR factorisation is the
default: $\oom{mn^{2}}$ flops, no squaring of conditioning, direct
production of $\mat{R}$ from which $\vect{x}$ is recovered by back
substitution. For symmetric positive-definite systems, Cholesky is
the default: half the work of LU, half the storage, and a clear
failure signal (a non-positive pivot detects a non-PD matrix
mid-factorisation). For matrix functions on non-symmetric matrices,
Schur is the default. SVD is the right tool when rank, conditioning,
optimal low-rank approximation, or the four fundamental subspaces
matter; for plain solves it is overkill.
