\chapter{Fluid dynamics: inviscid}
\label{vol03:ch11}

\epigraph{Things which are not yet known, are not on that account
to be confounded with things which are not. Pressure produced in
fluids by external causes is communicated equally in every
direction; and the velocity of efflux of a fluid at any opening is
that of a heavy body falling through a height equal to the height
of the fluid above the opening.}{Daniel Bernoulli, \emph{
Hydrodynamica} (1738), Section XII, in the Carmody and Kobus
translation \cite[pp.~244--245]{hist:bernoulli1738}.}

\chapmeta{Half-life: durable. Archetypes: balance, transport. Exercise target: 39. Project track: physical build (Pitot tube on a household fan). Hazard class: low. Reader path default: standard, with sections explicitly tagged. Named cases: Tacoma Narrows aeroelastic flutter (1940, cross-referenced to Chapter 13); Pitot-icing common-mode failure (Air France 447, cross-referenced to Chapter 12).}

\noindent
The previous chapter treated the fluid at rest: pressure as a
scalar field, hydrostatics as a special case of mechanical
equilibrium. This chapter sets the fluid in motion under the
assumption of zero viscosity. The inviscid idealisation cuts away
the thin boundary-layer shear that dominates wall drag, leaving the
bulk-flow physics that governs lift, pressure recovery, jet
trajectories, and free-surface waves. The reader leaves able to
write the continuity equation for any control volume, derive the
Euler momentum equation along a streamline, apply Bernoulli's
equation with full awareness of its assumptions, and recognise the
configurations in which a Bernoulli calculation gives the right
answer to within a few percent and those in which the same
calculation is qualitatively wrong.

\section{Continuity: mass conservation in a flowing fluid}\pathtag{core}

A flowing fluid carries mass, momentum, and energy through space.
The first conservation law is mass. Pick a fixed region of space,
a \engterm{control volume} $\mathcal{V}$ with boundary
$\partial\mathcal{V}$, and ask: how does the mass inside the
control volume change with time? The answer is the integral
continuity equation:
\[
\frac{d}{dt}\int_{\mathcal{V}} \rho \,dV
+ \oint_{\partial\mathcal{V}} \rho \vec{u} \cdot \hat{n}\,dA = 0,
\]
where $\rho$ is the local fluid density, $\vec{u}$ the local fluid
velocity, and $\hat{n}$ the outward unit normal on the boundary.
The first term is the rate of change of mass stored inside; the
second is the net mass flux leaving across the boundary. Their sum
is zero because mass is neither created nor destroyed inside the
control volume.

Applying the divergence theorem to the boundary integral and
collecting both terms under a single volume integral, the integrand
must vanish pointwise (since the control volume was arbitrary):
\[
\frac{\partial \rho}{\partial t} + \nabla\cdot(\rho\vec{u}) = 0.
\]
This is the \engterm{differential continuity equation}, valid at
every point of the flow at every instant. The two forms (integral
and differential) carry the same physical content; the integral
form is the working tool for a control-volume calculation around a
pipe, a tank, or an engine; the differential form is the working
tool for a partial-differential-equation analysis of the flow
field.

The divergence-theorem step deserves the direct construction, because
the same face-by-face bookkeeping returns when we build the momentum
equation. Take a small fixed rectangular box of sides $dx$, $dy$, $dz$
centred on the point $(x,y,z)$. The mass flux entering the left
$x$-face is the density times the normal velocity times the face area,
evaluated at $x - dx/2$,
\[
\left(\rho u_x\right)\Big|_{x-dx/2}\,dy\,dz
= \left(\rho u_x - \frac{\partial(\rho u_x)}{\partial x}\frac{dx}{2}
\right)dy\,dz,
\]
expanding to first order in the small width. The flux leaving the
right $x$-face, at $x + dx/2$, carries the opposite-sign Taylor
correction, and the two combine to a net $x$-outflow of
$\partial(\rho u_x)/\partial x\,\,dx\,dy\,dz$. The $y$- and $z$-faces
contribute the analogous terms, so the total net mass outflow from the
box is
\[
\left[\frac{\partial(\rho u_x)}{\partial x}
+ \frac{\partial(\rho u_y)}{\partial y}
+ \frac{\partial(\rho u_z)}{\partial z}\right]dx\,dy\,dz
= \nabla\cdot(\rho\vec u)\,dV.
\]
That net outflow must be supplied by the fall in stored mass,
$-\partial\rho/\partial t\,\,dV$, and equating the two and cancelling
the common volume recovers the differential equation directly, with no
appeal to a theorem. The divergence operator is the per-unit-volume net
outflow, read off the six faces; carrying the same construction with
the momentum flux in place of the mass flux is the whole of the
derivation in the next section.

For an \engterm{incompressible} fluid, $\rho$ is constant in space
and time, and the differential continuity equation reduces to
\[
\nabla\cdot\vec{u} = 0.
\]
The velocity field of an incompressible flow is solenoidal: it has
zero divergence at every point. Liquids satisfy the incompressible
condition to a very good approximation under nearly all engineering
conditions; the bulk modulus of water is about $2.2\,\text{GPa}$,
giving a density change of about $0.05\,\%$ under a $1\,\text{MPa}$
pressure rise. Gases satisfy the incompressible condition only at
flow speeds well below the local speed of sound, the criterion of
chapter 13 of this volume; at speeds above about $0.3 M$, density
variations of more than $5\,\%$ make incompressible analysis
inaccurate.

For a steady one-dimensional flow through a pipe or channel, the
integral continuity equation simplifies further. With no time
variation and with the inlet and outlet streams normal to the
control surface,
\[
\rho_1 A_1 u_1 = \rho_2 A_2 u_2,
\]
where subscripts $1$ and $2$ label the inlet and outlet
cross-sections, $A$ is the cross-sectional area, and $u$ is the
flow speed averaged over the cross-section. For an incompressible
fluid, $\rho_1 = \rho_2$ and the relation reduces to
\[
A_1 u_1 = A_2 u_2,
\]
the \engterm{conservation of volume flow rate} $Q = A u$. A pipe
that narrows from $100\,\si{\milli\meter}$ to $50\,\si{\milli\meter}$
diameter accelerates the flow by a factor of four (the area ratio
is the square of the diameter ratio). The acceleration is purely
geometric; no force has been invoked yet. The next section supplies
the force.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-control-volume.tex}

Figure~\ref{fig:vol03ch11:control-volume} draws the bookkeeping on a
contracting duct: mass enters one face, leaves the other, and (in
steady flow) nothing accumulates between. The reader who can sketch
this control volume and write the two face fluxes has the whole of
mass conservation in hand; everything later in the chapter rests on
closing this balance first.

\begin{principle}[Continuity is a bookkeeping identity]
Mass entering a control volume per unit time equals mass leaving per
unit time plus mass accumulating inside. The relation is exact for
any control volume in any flow; it requires no assumption about the
constitutive behaviour of the fluid. For an incompressible flow
through a pipe, the velocity scales inversely with the
cross-sectional area; for a compressible flow, the density carries
part of the variation, and the velocity scales with $1/(\rho A)$.
Every fluid-mechanics calculation starts by closing the mass
balance.
\end{principle}

\section{Euler's equation: momentum conservation in an inviscid fluid}\pathtag{core}

Newton's second law applied to a fluid particle of mass $\rho\,dV$
gives
\[
\rho\,dV\,\frac{D\vec{u}}{Dt} = \vec{F}_{\text{on the particle}},
\]
where $D/Dt = \partial/\partial t + \vec{u}\cdot\nabla$ is the
\engterm{material derivative}, the time rate of change following
the fluid particle (rather than at a fixed point in space). The
material derivative carries the convective acceleration
$(\vec{u}\cdot\nabla)\vec{u}$, which records the change in velocity
the particle experiences because it moves into a region where the
velocity is different.

The forces on the particle are of two kinds: \engterm{body forces}
(gravity, electromagnetic forces) that act on the volume, and
\engterm{surface forces} (pressure, viscous stress) that act on the
particle's surface. For an \engterm{inviscid fluid}, the surface
force reduces to the pressure alone: the surrounding fluid pushes
on the particle with a force per unit area equal to the local
pressure, directed inward along the surface normal. The net surface
force per unit volume on the particle is $-\nabla p$.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-differential-element.tex}

The $-\nabla p$ surface force is the face-pressure bookkeeping of
Figure~\ref{fig:vol03ch11:differential-element}, worked through one
coordinate at a time. Pressure on the left $x$-face pushes the box to
the right with force $\big(p - (\partial p/\partial x)(dx/2)\big)dy\,dz$;
pressure on the right $x$-face pushes it to the left with force
$\big(p + (\partial p/\partial x)(dx/2)\big)dy\,dz$. The two centre
terms cancel and the net $x$-force is $-(\partial p/\partial x)\,
dx\,dy\,dz$. The $y$- and $z$-faces give the matching components, so
the net pressure force on the box is $-\nabla p\,\,dV$, and dividing by
the volume leaves the force per unit volume $-\nabla p$. The minus sign
is the physical content: a fluid particle is pushed from high pressure
toward low pressure, down the gradient. Adding the gravitational body
force $\rho \vec{g}$ gives
\[
\rho\,\frac{D\vec{u}}{Dt} = -\nabla p + \rho\vec{g}.
\]
This is the \engterm{Euler equation}, derived by Leonhard Euler in
his 1755 \emph{Principes g\'en\'eraux du mouvement des fluides}.
The Euler equation is the inviscid limit of the Navier-Stokes
equation (chapter 12 of this volume); it is the momentum-balance
law for a fluid in which viscosity is neglected.

Expanding the material derivative,
\[
\frac{\partial \vec{u}}{\partial t}
+ (\vec{u}\cdot\nabla)\vec{u}
= -\frac{1}{\rho}\nabla p + \vec{g}.
\]
The first term on the left is the \engterm{local acceleration} (the
unsteadiness of the flow at a fixed point); the second is the
\engterm{convective acceleration} (the spatial variation the
particle samples as it moves). The convective term is nonlinear in
$\vec{u}$; this nonlinearity is the source of most of fluid
dynamics' analytical difficulty, including turbulence (chapter 12)
and shock waves (chapter 13).

For a \engterm{steady flow}, $\partial \vec{u}/\partial t = 0$, and
the Euler equation becomes a statement about the spatial structure
of the flow field. Taking the dot product of the steady Euler
equation with a unit tangent $\hat{s}$ along a streamline,
\[
(\vec{u}\cdot\nabla)\vec{u}\cdot\hat{s}
= u\,\frac{\partial u}{\partial s}
= \frac{1}{2}\frac{\partial(u^2)}{\partial s},
\]
where $s$ is arc length along the streamline. The pressure term
becomes $-(1/\rho)\,\partial p/\partial s$, and the gravity term
becomes $-g\,\partial z/\partial s$ (with $z$ measured against
gravity). Multiplying through by $\rho$ and integrating along the
streamline gives Bernoulli's equation, the subject of the next
section.

The Euler equation also has a \engterm{normal} form, taken
perpendicular to the streamline. For a streamline with local radius
of curvature $R_c$, the centripetal acceleration is $u^2/R_c$ toward
the centre of curvature, and the normal component of Euler's
equation gives
\[
\frac{\rho u^2}{R_c} = -\frac{\partial p}{\partial n}
+ \rho g_n,
\]
where $n$ is the inward normal coordinate and $g_n$ is the gravity
component along it. The relation says: a streamline curves because
the pressure gradient across it (plus any gravity component) pushes
the fluid particle inward. The pressure is higher on the outside
of a curved streamline than on the inside. This is the working
intuition behind every aerodynamic surface: a curved streamline
over the top of an airfoil corresponds to a pressure low on the
upper surface relative to the freestream, and this pressure low
times the upper surface area is what lifts the wing.

\subsection{The integral momentum theorem}

The Euler equation is the pointwise momentum balance. Integrating it
over a fixed control volume, or applying Newton's second law directly
to the mass inside, gives the \engterm{integral momentum theorem},
the working tool whenever the forces on a device matter more than the
detailed flow field:
\[
\sum \vec{F}
= \frac{d}{dt}\int_{\mathcal{V}} \rho\vec{u}\,dV
+ \oint_{\partial\mathcal{V}} \rho\vec{u}\,(\vec{u}\cdot\hat{n})\,dA.
\]
The left side collects every force on the control volume: pressure
forces on the inlet and outlet faces, the reaction force from any
solid surface cut by the boundary, gravity, and any external
restraint. The right side is the rate of change of stored momentum
plus the net momentum flux carried out across the boundary. For
steady flow with one inlet and one outlet, the relation reduces to
\[
\sum \vec{F} = \dot{m}\,(\vec{u}_2 - \vec{u}_1),
\quad \dot{m} = \rho A u,
\]
the form a working engineer uses to size the anchor bolts on a pipe
bend, the thrust of a jet, or the force a water jet exerts on a
turbine bucket. The momentum theorem makes no inviscid assumption: it
holds across a hydraulic jump, through a separated wake, and inside a
turbomachine, the configurations where Bernoulli does not. That
contrast, energy not conserved but momentum always conserved, is the
diagnostic backbone of the failure section.

\subsection{The angular-momentum theorem}

Taking the moment of the momentum balance about a fixed axis turns the
force theorem into a torque theorem, the working tool for everything
that spins: pump and turbine rotors, the recoil of a sprinkler, the
torque on a pipe manifold. Forming the cross product of the position
vector $\vec r$ (from the axis to each point) with the momentum balance
and integrating over the control volume gives the
\engterm{angular-momentum theorem}
\[
\sum \vec{M}
= \frac{d}{dt}\int_{\mathcal{V}} \vec r\times\rho\vec{u}\,dV
+ \oint_{\partial\mathcal{V}}
(\vec r\times\rho\vec{u})\,(\vec{u}\cdot\hat{n})\,dA,
\]
where $\sum\vec M$ is the net torque of all forces about the axis. The
first term on the right is the rate of change of stored angular
momentum; the second is the net flux of angular momentum carried out
across the boundary. For steady flow with one inlet and one outlet the
stored term vanishes and, taking the scalar component about a fixed
axis with $r_t$ the moment arm of the tangential velocity $u_t$ at each
port, the theorem reduces to
\[
M = \dot m\,(r_{t,2}\,u_{t,2} - r_{t,1}\,u_{t,1}),
\]
the \engterm{Euler turbomachine equation} in its rawest form: the
torque on the rotor equals the mass flow times the change in the
product of radius and swirl velocity between inlet and outlet. A pump
adds swirl and so demands a driving torque; a turbine removes swirl and
so delivers one; a sprinkler ejects swirl it never had at inlet and so
recoils. The angular-momentum theorem inherits the linear theorem's
robustness, it makes no inviscid assumption, so it sizes the shaft
torque of a real machine where Bernoulli could not follow the flow
through the blading.

\begin{mastery}
\textbf{The Reynolds transport theorem: one identity behind every
control-volume law.} The continuity equation, the momentum theorem, and
the angular-momentum theorem are three readings of a single kinematic
identity, and naming it once saves rederiving the conversion each time.
Let $B$ be any extensive property of the fluid (mass, momentum, angular
momentum, energy) and $b = dB/dm$ its value per unit mass. The rate of
change of $B$ for the \engterm{system}, the fixed collection of fluid
particles that happens to occupy the control volume at the instant in
question, relates to what happens in the fixed \engterm{control volume}
by
\[
\frac{dB_\text{sys}}{dt}
= \frac{d}{dt}\int_{\mathcal V}\rho b\,dV
+ \oint_{\partial\mathcal V}\rho b\,(\vec u\cdot\hat n)\,dA.
\]
The system derivative on the left is where a physical law applies:
mass is conserved ($dB_\text{sys}/dt = 0$ with $b = 1$), momentum
obeys Newton's second law ($dB_\text{sys}/dt = \sum\vec F$ with
$b = \vec u$), angular momentum obeys the moment law
($dB_\text{sys}/dt = \sum\vec M$ with $b = \vec r\times\vec u$). The
two terms on the right are the same pair every control-volume balance
carries: the rate of accumulation inside the fixed volume, and the net
flux out across its boundary, the flux always written as the per-mass
property $b$ times the mass flux $\rho(\vec u\cdot\hat n)\,dA$.
Substituting the three choices of $b$ regenerates the three theorems of
this chapter without any fresh derivation; substituting $b = e +
p/\rho + \tfrac12 u^2 + gz$ regenerates the steady-flow energy equation
the failure section calls on for turbomachinery. The theorem is the
bridge between the Lagrangian statement of a physical law, which holds
for a fixed lump of matter, and the Eulerian control volume, which the
engineer can actually draw around a pump or a pipe bend. Every balance
in this chapter is one substitution into this identity, and recognising
that collapses what looks like four separate bookkeeping rules into one.
\end{mastery}

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-momentum-bend.tex}

\paragraph{Worked example: force on a $90^\circ$ pipe bend.} Water
($\rho = 1000\,\si{\kilogram\per\meter\cubed}$) flows at
$3\,\si{\meter\per\second}$ through a $100\,\si{\milli\meter}$
diameter bend that turns the flow from horizontal to vertical
(Figure~\ref{fig:vol03ch11:momentum-bend}). The gauge pressure at the
inlet is $200\,\text{kPa}$; for an inviscid bend of constant area the
speed is unchanged and, taking the bend short enough to neglect the
elevation change, the pressure is unchanged through the bend, so
$u_1 = u_2 = 3\,\si{\meter\per\second}$ and $p_2 = p_1 =
200\,\text{kPa}$. The area is $A = \pi(0.05)^2 = 7.85\times
10^{-3}\,\si{\meter}^2$ and the mass flow is $\dot{m} = \rho A u =
1000\times 7.85\times 10^{-3}\times 3 = 23.6\,\si{\kilogram\per
\second}$. Take $x$ along the inlet flow and $y$ along the outlet
flow. The $x$-momentum balance reads $p_1 A + R_x = \dot{m}(0 - u_1)$,
giving $R_x = -p_1 A - \dot{m}u_1 = -(200\,000)(7.85\times 10^{-3}) -
(23.6)(3) = -1571 - 71 = -1642\,\si{\newton}$. The $y$-balance reads
$-p_2 A + R_y = \dot{m}(u_2 - 0)$, giving $R_y = p_2 A + \dot{m}u_2 =
1571 + 71 = 1642\,\si{\newton}$. The force the bolts must carry is the
reaction $-\vec{R}$, of magnitude $\sqrt{R_x^2 + R_y^2} =
2320\,\si{\newton}$ directed back along the bisector of the bend. The
pressure term dominates the momentum term by a factor of about
twenty; a designer who sizes the restraint from the momentum change
alone underestimates the load by that factor.

\section{Bernoulli's equation: the energy identity along a streamline}\pathtag{core}

Integrating the steady Euler equation along a streamline, with
$\rho$ constant (incompressible flow),
\[
\int_1^2 \left(u\,\frac{\partial u}{\partial s}
+ \frac{1}{\rho}\frac{\partial p}{\partial s}
+ g\frac{\partial z}{\partial s}\right) ds = 0,
\]
which yields the \engterm{Bernoulli equation}
\[
\frac{1}{2}u^2 + \frac{p}{\rho} + gz = \text{constant along a
streamline}.
\]
In Bernoulli's original presentation in the \emph{Hydrodynamica} of
1738, the relation was written in terms of velocity heads and
pressure heads
\cite[Section XII]{hist:bernoulli1738}; the form above is the
modern energy-density version.

The streamline integration just carried out is the first of two routes
to the same constant, and the two differ in what they assume and in how
widely the constant applies. The streamline route used only
steadiness, incompressibility, and the inviscid balance; it makes no
assumption about rotation, and it pays for that generality with a
constant that holds along one streamline and may differ on the next.
The second route assumes the flow is also irrotational and earns a
constant good across the whole field. For an irrotational flow the
velocity is a gradient, $\vec u = \nabla\phi$, and the convective term
rewrites with the Lamb identity
\[
(\vec u\cdot\nabla)\vec u
= \nabla\!\left(\tfrac12 u^2\right) - \vec u\times(\nabla\times\vec u)
= \nabla\!\left(\tfrac12 u^2\right),
\]
the curl term dropping because $\nabla\times\nabla\phi = 0$. The full
Euler equation, with the local-acceleration term carried, becomes a
single gradient set to zero,
\[
\nabla\!\left(\frac{\partial\phi}{\partial t} + \tfrac12 u^2
+ \frac{p}{\rho} + gz\right) = 0,
\]
so the bracket is at most a function of time, the same on every
streamline at a given instant. For steady irrotational flow the
$\partial\phi/\partial t$ term vanishes and the bracket is a single
constant throughout the field, not merely along each streamline. The
two routes meet for steady irrotational incompressible flow, where the
Bernoulli constant is one number everywhere; they part for a rotational
flow, where the streamline route still holds streamline by streamline
but the field-wide statement fails, the distinction Crocco's theorem
makes precise later in the chapter and the failure section turns into a
working warning.

The relation says: along a streamline, the sum of kinetic energy
per unit mass, pressure-work per unit mass, and gravitational
potential energy per unit mass is conserved. The Bernoulli sum is
the mechanical energy density of the fluid particle as it travels
along its streamline. Energy can convert between kinetic and
pressure forms (the fluid speeds up as the pressure drops, and
vice versa), but the total along any one streamline is the same at
every point.

Multiplying through by $\rho$ gives the \engterm{pressure form} of
Bernoulli's equation:
\[
\frac{1}{2}\rho u^2 + p + \rho g z = p_0 = \text{const.\ along a
streamline}.
\]
The constant $p_0$ is the \engterm{total pressure} or \engterm{
stagnation pressure}: the pressure that a fluid particle would have
if brought to rest adiabatically and reversibly at the same elevation.
The term $\tfrac{1}{2}\rho u^2$ is the \engterm{dynamic pressure}
$q$, and $p$ is the \engterm{static pressure}.

Dividing through by $\rho g$ gives the \engterm{head form}, in
which each term has units of length:
\[
\frac{u^2}{2g} + \frac{p}{\rho g} + z = H = \text{const.\ along a
streamline},
\]
with $H$ the \engterm{total head}. The head form is the working
language of hydraulic engineering; canal slopes, pump duty curves,
and dam spillway calculations are routinely written in heads of
water.

The assumptions on which Bernoulli's equation rests are five, and
the working engineer carries the list in active memory:

\begin{itemize}[leftmargin=*]
  \item \engterm{Steady flow}: $\partial\vec{u}/\partial t = 0$
    at every point. Pulsating flows, starting transients, and
    waves violate this assumption.
  \item \engterm{Incompressible}: $\rho$ constant along the
    streamline. Gases at Mach $> 0.3$ violate this; the
    corresponding extension is the compressible Bernoulli equation
    of chapter 13 of this volume.
  \item \engterm{Inviscid}: zero viscous stress. Real fluids in
    real geometries develop boundary layers (chapter 12 of this
    volume) where viscous effects dominate; Bernoulli does not
    apply inside the boundary layer.
  \item \engterm{Along a streamline}: the constant differs between
    different streamlines unless the flow is also irrotational
    (next section).
  \item \engterm{No heat addition, no shaft work, no separation}:
    the streamline must run from inlet to outlet through a
    well-defined flow without recirculation or energy exchange.
\end{itemize}

When all five assumptions are met, Bernoulli's equation is exact.
When one or more are violated, the equation is either approximate
(the most common case) or qualitatively wrong (the case studied in
the failure section of this chapter).

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-bernoulli-streamline.tex}

Figure~\ref{fig:vol03ch11:bernoulli-streamline} makes the trade
visible: along a streamline through a throat, the velocity head and
the pressure head exchange magnitude while their sum, the total head,
holds flat. The flat total head is the geometric content of
Bernoulli's equation. A real duct with friction would show the total
head sloping gently downward; reading the slope of the total-head
line is how a hydraulic engineer separates the inviscid prediction
from the viscous loss.

\begin{principle}[Bernoulli is conservation of mechanical energy along a streamline]
For a steady, incompressible, inviscid flow, the quantity
$\tfrac{1}{2}\rho u^2 + p + \rho g z$ is conserved along any
streamline. The pressure drops where the fluid speeds up and rises
where the fluid slows; the elevation enters as a gravitational
potential. The relation is the working tool for jet velocity,
Pitot-tube anemometry, Venturi flow metering, pump performance, and
free-surface flow analysis.
\end{principle}

\begin{mastery}
\textbf{The hydraulic and energy grade lines: Bernoulli drawn as two
elevations.} The head form of Bernoulli invites a picture that a
hydraulic engineer reads a whole system from, and setting it down once
turns the algebra into a drawing. Plot, above an arbitrary datum, two
lines along the length of a pipe or channel. The \engterm{energy grade
line} (EGL) is the total head $z + p/(\rho g) + u^2/(2g)$, the height to
which the fluid's mechanical energy would raise it. The \engterm{
hydraulic grade line} (HGL) is the piezometric head $z + p/(\rho g)$
alone, the level a static tapping standpipe would show, sitting below
the EGL by exactly the velocity head $u^2/(2g)$. Four readings follow
at a glance. For an ideal Bernoulli flow the EGL is horizontal, because
the total head is conserved; any downward slope of the EGL is a head
loss, and its steepness is the loss rate, so a real pipe's EGL tilts
gently down and drops a step at every fitting. The gap between the EGL
and the HGL is the velocity head, so where a pipe narrows the HGL dives
toward the EGL as the flow speeds up, and where it widens the HGL rises
back. Where the HGL falls below the pipe itself the gauge pressure is
negative, a suction the walls must not collapse under and a place the
flow may cavitate or draw air at a joint, which is how a designer spots
a siphon crest or a pump-suction trouble on the drawing before it is
built. A pump adds a vertical step up to both lines (it raises the total
head), a turbine takes a step down, and a free jet leaves with its HGL
at the jet surface and its EGL a velocity head above. The EGL and HGL
are Bernoulli's equation made into a profile the eye reads, and the
metering-station and penstock case studies later in the chapter both
carry one; a reader who can sketch the two lines for a proposed layout
has checked the energy balance of the whole system without writing a
single number. The construction is developed at length in White's
treatment of pipe systems
\cite[chap.~6]{text:white-fluid-mechanics}.
\end{mastery}

\subsection{Three standard applications}

\engterm{Torricelli's law}. A tank of liquid with a small hole at
depth $h$ below the free surface drains through the hole at speed
\[
u = \sqrt{2 g h},
\]
the same speed a body would acquire falling freely through height
$h$. The result follows from Bernoulli along a streamline from the
free surface (where $u_1 \approx 0$ by the assumption that the tank
is much larger than the hole) to the jet (where $p_2 \approx p_1$
because both connect to the atmosphere), with $z_1 - z_2 = h$. The
result was first stated by Evangelista Torricelli in 1644 and is
sometimes used as the starting axiom of fluid dynamics; Bernoulli's
derivation places it within a wider conservation principle. The
efflux speed $\sqrt{2gh}$ is the instantaneous value; when the tank
drains, $h$ falls and the problem becomes unsteady. Volume
conservation, $A_s\,dh/dt = -C_d A_o\sqrt{2gh}$, integrates to a drain
time $t = (A_s/C_d A_o)\sqrt{2h_0/g}$. The script
\path{code/tank_drain.py} integrates this ODE and checks it against
the closed form, writing the depth history to
\path{data/tank_drain_history.csv}; for a $25\,\si{\meter}\times
10\,\si{\meter}$ pool $2\,\si{\meter}$ deep draining through a
$100\,\si{\milli\meter}$ outlet, the two agree at about nine hours.

\engterm{Pitot tube}. A Pitot tube has a forward-facing opening
that registers the stagnation pressure $p_0$ and a static-pressure
port that registers $p$. The difference is the dynamic pressure
$q = p_0 - p = \tfrac{1}{2}\rho u^2$, from which
\[
u = \sqrt{\frac{2(p_0 - p)}{\rho}}.
\]
Pitot-static probes on aircraft, on wind-tunnel test sections, and
on ship hulls all rest on this relation. The probe's calibration
adjustment is typically a small geometric correction; the working
physics is Bernoulli.

\engterm{Venturi meter}. A pipe with a contraction from area $A_1$
to area $A_2$ and back accelerates the flow through the throat
and decelerates it back to the original speed. Continuity gives
$u_2 = u_1 (A_1/A_2)$; Bernoulli along the centreline gives
$p_1 - p_2 = \tfrac{1}{2}\rho(u_2^2 - u_1^2)$. Combining,
\[
u_1 = \sqrt{\frac{2(p_1 - p_2)}{\rho [(A_1/A_2)^2 - 1]}}.
\]
The flow rate $Q = A_1 u_1$ follows. The Venturi meter measures
flow rate by measuring a pressure differential; the relation is
calibrated with a discharge coefficient $C_d \approx 0.97$ that
accounts for small viscous losses (the topic of chapter 12 of this
volume). Orifice plates use the same physics with a discharge
coefficient typically in the range $0.60$ to $0.65$.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-venturi-pitot.tex}

Figure~\ref{fig:vol03ch11:venturi-pitot} sets the Venturi meter and
the Pitot-static probe side by side; both turn a velocity measurement
into a pressure measurement, and both fail in the same way, through a
static port that does not read the true static pressure. The script
\path{code/venturi_flow.py} computes the flow rate from a measured
differential and sweeps the differential pressure to produce the
calibration table \path{data/venturi_calibration.csv}; for the
$100/40\,\si{\milli\meter}$ meter of the chapter exercise a
$30\,\text{kPa}$ differential corresponds to a throat speed of
$7.85\,\si{\meter\per\second}$ and a flow rate of $9.57\,\text{L/s}$.

\section{Streamlines, streaklines, pathlines}\pathtag{standard}

A flow field is a vector field $\vec{u}(\vec{x}, t)$. Three
geometric constructs describe the field's structure:

\engterm{Streamline}. A curve that is everywhere tangent to the
velocity field at a given instant. The differential equation of a
streamline at time $t_0$ is
\[
\frac{dx}{u_x(\vec{x}, t_0)} = \frac{dy}{u_y(\vec{x}, t_0)} =
\frac{dz}{u_z(\vec{x}, t_0)}.
\]
A streamline is a snapshot of the flow direction; streamlines
cannot cross (except at stagnation points, where the velocity is
zero).

\engterm{Pathline}. The trajectory of a specific fluid particle as
it moves through the flow over time. The differential equation is
\[
\frac{d\vec{x}_p}{dt} = \vec{u}(\vec{x}_p, t), \quad \vec{x}_p(0) =
\vec{x}_0,
\]
giving the particle's position $\vec{x}_p(t)$ as a function of time
from its initial position $\vec{x}_0$.

\engterm{Streakline}. The locus, at one instant, of all fluid
particles that have passed through a given point in the flow. A
plume of dye injected at a fixed location and photographed at one
instant traces a streakline.

For a \engterm{steady flow}, all three coincide: streamlines,
pathlines, and streaklines are the same curves in space, because
the flow field at any later time is the same as the flow field
at the initial time. For an unsteady flow they diverge, and the
distinction matters for interpreting flow visualisation.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-streamline-types.tex}

Figure~\ref{fig:vol03ch11:streamline-types} shows the three curves
pulling apart in a flow with steady drift and an oscillating
cross-component. A wind-tunnel smoke trace is a streakline, a tracked
particle is a pathline, and a computed instantaneous tangent field is
a set of streamlines; reading a smoke photograph as if it showed
streamlines is a standard error in unsteady flow, and it is one the
reader should be able to catch.

\engterm{Stream function (two-dimensional)}. For a
two-dimensional incompressible flow, the continuity equation
$\partial u_x/\partial x + \partial u_y/\partial y = 0$ is
satisfied identically by introducing a scalar \engterm{stream
function} $\psi(x, y)$ such that
\[
u_x = \frac{\partial \psi}{\partial y}, \quad
u_y = -\frac{\partial \psi}{\partial x}.
\]
Lines of constant $\psi$ are streamlines: along a curve where
$\psi$ does not change, the velocity component normal to the curve
is zero, so the curve is everywhere tangent to the flow. The flow
rate between two streamlines is the difference of their stream
function values, $Q = \psi_2 - \psi_1$. The stream function is the
working tool for two-dimensional incompressible flow analysis; it
turns a vector-field problem into a scalar-field problem with one
fewer dependent variable.

For an \engterm{axisymmetric flow} (cylindrical coordinates $r,
\theta, z$ with no $\theta$-dependence and $u_\theta = 0$), the
Stokes stream function $\psi(r, z)$ satisfies
\[
u_z = \frac{1}{r}\frac{\partial \psi}{\partial r}, \quad
u_r = -\frac{1}{r}\frac{\partial \psi}{\partial z},
\]
and again streamlines are lines of constant $\psi$.

The stream function reads off a flow rate without an integration. For
the cylinder-in-uniform-flow solution introduced later in this chapter,
$\psi = U_\infty r\sin\theta\,(1 - R^2/r^2)$, the difference in $\psi$
between two points is the volume flow per unit depth crossing any line
joining them. Between the cylinder surface ($\psi = 0$) and a point one
radius above the top ($r = 2R$, $\theta = \pi/2$, where $\psi =
U_\infty\,2R\,(1 - \tfrac14) = 1.5\,U_\infty R$), the flow per unit
depth passing through that gap is $1.5\,U_\infty R$, larger than the
$1.0\,U_\infty R$ that would cross the same height in the undisturbed
freestream because the cylinder crowds the streamlines together over
its crown. The crowding is the geometric form of the speed-up that
Bernoulli converts into the surface suction; reading streamline spacing
is reading speed, and reading speed is reading pressure.

\section{Vorticity and circulation}\pathtag{core}

The local rotation rate of a fluid particle is the \engterm{
vorticity}
\[
\vec{\omega} = \nabla\times\vec{u}.
\]
Vorticity is twice the angular velocity of the fluid particle
about its centre. A flow with zero vorticity at every point is
\engterm{irrotational}; a flow with non-zero vorticity is
\engterm{rotational}.

For an irrotational flow, the velocity field is the gradient of a
scalar potential:
\[
\vec{u} = \nabla\phi,
\]
where $\phi$ is the \engterm{velocity potential}. The combination
of incompressibility ($\nabla\cdot\vec{u} = 0$) and
irrotationality ($\vec{u} = \nabla\phi$) gives
\[
\nabla^2\phi = 0,
\]
Laplace's equation. Solutions of Laplace's equation are
\engterm{harmonic functions}, the same class that governs
electrostatic potentials, steady heat conduction in a homogeneous
solid, and gravitational potentials in source-free regions. The
mathematical correspondence is the reason that techniques developed
for one field translate directly to the others: the method of
images for electrostatics carries over to the method of images for
potential flow; complex-variable methods that map heat-conduction
problems map fluid-flow problems with the same conformal-mapping
machinery.

The \engterm{circulation} around a closed curve $C$ is the line
integral
\[
\Gamma = \oint_C \vec{u}\cdot d\vec{\ell}.
\]
By Stokes's theorem, $\Gamma$ equals the flux of vorticity through
any surface bounded by $C$:
\[
\Gamma = \int_S \vec{\omega}\cdot \hat{n}\,dA.
\]
For an irrotational flow in a simply connected region, the
circulation around any closed curve is zero. For an irrotational
flow in a multiply connected region (one with a hole through it,
for example a curve enclosing an airfoil), the circulation around
a curve enclosing the hole can be non-zero; the value is the same
for any curve that encloses the same set of holes, by the no-vorticity
condition in the rest of the region. The circulation around the
airfoil is what produces its lift.

\engterm{Kelvin's circulation theorem} (Lord Kelvin, 1869): for an
inviscid, incompressible (or barotropic) flow under conservative
body forces, the circulation around a material curve (a curve made
up of the same fluid particles at every instant) is conserved as
the curve moves with the flow:
\[
\frac{D\Gamma}{Dt} = 0.
\]
The theorem has a striking consequence: if a flow starts from rest
(zero circulation everywhere), it remains everywhere irrotational
unless viscosity or compressibility or non-conservative forces
introduce vorticity. In particular, an airfoil placed in a uniform
freestream and inviscid flow cannot develop circulation. The lift
of a real airfoil exists because viscosity, confined to a thin
boundary layer at the trailing edge, sheds a \engterm{starting
vortex} of opposite circulation as the wing accelerates from rest,
leaving the wing with a bound circulation equal and opposite to the
shed circulation, by Kelvin's theorem applied to the original
material curve that surrounded both the wing and the eventual shed
vortex.

\engterm{Helmholtz's vortex theorems} (Hermann von Helmholtz,
1858):

\begin{itemize}[leftmargin=*]
  \item Vortex lines (lines tangent to $\vec{\omega}$ at every
    point) move with the fluid in an inviscid barotropic flow.
  \item A vortex tube has constant strength along its length;
    vortex tubes cannot end inside a fluid, they must terminate
    on a boundary or close on themselves.
  \item The strength of a vortex tube (the circulation around a
    curve enclosing the tube once) is conserved in time.
\end{itemize}

The three theorems together govern the topology of vorticity in
inviscid flow. Smoke rings, tornado vortices, and the trailing
vortices of an aircraft wing all obey them; the wing vortices
extend downstream of the aircraft as a vortex pair, eventually
descending and merging or dissipating into atmospheric turbulence,
because the trailing vortex tubes must terminate on themselves
(no fluid-mechanical reason allows them to simply end).

\begin{mastery}
\textbf{The vorticity transport equation.} Taking the curl of the
Euler equation removes the pressure term (the curl of a gradient
vanishes) and gives the evolution of vorticity directly. For an
incompressible inviscid flow under conservative body forces,
\[
\frac{D\vec{\omega}}{Dt}
= (\vec{\omega}\cdot\nabla)\vec{u}.
\]
The left side is the rate of change of vorticity following a fluid
particle; the right side is the \engterm{vortex-stretching} term. It
has no analogue in two dimensions: when $\vec{\omega}$ and $\nabla$
lie in the same plane and $\vec{u}$ has no out-of-plane component,
$(\vec{\omega}\cdot\nabla)\vec{u} = 0$, so two-dimensional inviscid
vorticity is simply carried along unchanged, $D\omega/Dt = 0$. In
three dimensions the stretching term lets a vortex tube intensify as
it is drawn out and thinned, conserving its circulation (Helmholtz)
while raising its local vorticity. This is the mechanism by which a
bath drain spins faster as it narrows, by which a tornado concentrates
weak background rotation into a violent core, and by which turbulent
energy cascades from large eddies to small. Kelvin's circulation
theorem and the vorticity transport equation are the same physics
seen integrally and differentially: circulation around a material
loop is conserved, and the vorticity threading the loop rearranges
itself by stretching and tilting as the loop deforms. Viscosity,
absent from this inviscid statement, is what diffuses vorticity
across streamlines and ultimately dissipates it; its omission here is
exactly why an inviscid flow started from rest can never generate the
bound circulation a real wing carries without the boundary-layer
shedding of a starting vortex.
\end{mastery}

\begin{mastery}
\textbf{Crocco's theorem: where vorticity comes from.} Bernoulli says
the total mechanical energy is constant along a streamline; the
question of whether it is the same constant on neighbouring streamlines
is what links the energy picture to the vorticity picture. Write the
steady Euler equation with the convective term in the Lamb form
$(\vec u\cdot\nabla)\vec u = \nabla(\tfrac12 u^2) - \vec u\times
\vec\omega$. For a steady flow under a conservative body force the
balance rearranges to
\[
\vec u\times\vec\omega = \nabla\!\left(\tfrac12 u^2 + \frac{p}{\rho}
+ gz\right) = \nabla B,
\]
where $B$ is the Bernoulli function, the quantity Bernoulli's equation
holds constant along a streamline. The cross product on the left is
perpendicular to $\vec u$, so $\vec u\cdot\nabla B = 0$ recovers the
streamwise statement at once: $B$ is constant along a streamline. The
new content is the component across streamlines. If $B$ varies from one
streamline to the next, the gradient $\nabla B$ is non-zero in the
cross-stream direction, and the equation forces $\vec u\times\vec\omega
\neq 0$: the flow must carry vorticity. A flow in which $B$ is the same
constant everywhere (a \engterm{homenergic} flow) and which is also
irrotational stays irrotational; a flow with a cross-stream gradient of
$B$ cannot be irrotational. For a compressible flow the same derivation,
carried with the enthalpy in place of $p/\rho$, becomes Crocco's
theorem in its usual form $T\nabla s = \nabla h_0 - \vec u\times
\vec\omega$, relating the entropy gradient, the total-enthalpy
gradient, and the vorticity. The reading that matters for this chapter
is the incompressible corollary: the Bernoulli constant differs
between streamlines exactly when the flow is rotational, which is why
the failure section's warning, that Bernoulli cannot be carried across
streamlines in a rotational flow, is not a separate rule but a direct
consequence of the momentum balance. A curved shock, which leaves
different streamlines with different stagnation states, is the standard
generator of the cross-stream gradient and therefore of downstream
vorticity, the bridge to the compressible flow of chapter 13 of this
volume.
\end{mastery}

\section{Potential flow: superposition of elementary solutions}\pathtag{standard}

\subsection{The potential, the stream function, and their orthogonality}

Two scalar fields carry the whole of a two-dimensional incompressible
irrotational flow, and the relation between them is worth setting down
before the superposition machinery uses both. The velocity potential
$\phi$ exists because the flow is irrotational ($\vec u = \nabla\phi$);
the stream function $\psi$ exists because the flow is incompressible
(continuity is satisfied identically by $u_x = \partial\psi/\partial y$,
$u_y = -\partial\psi/\partial x$). Writing the velocity components both
ways equates them:
\[
u_x = \frac{\partial\phi}{\partial x} = \frac{\partial\psi}{\partial y},
\qquad
u_y = \frac{\partial\phi}{\partial y} = -\frac{\partial\psi}{\partial x}.
\]
These are the \engterm{Cauchy-Riemann equations}, the same pair that
makes a complex function analytic. Three consequences follow at once.
First, each field is harmonic: differentiating the first relation by
$x$ and the second by $y$ and adding gives $\nabla^2\phi = 0$, and the
reverse pairing gives $\nabla^2\psi = 0$, so both the potential and the
stream function satisfy Laplace's equation. Second, the level sets are
orthogonal. The gradient $\nabla\phi = (u_x, u_y)$ points along the
flow, while $\nabla\psi = (-u_y, u_x)$ is the same vector turned a
right angle, so $\nabla\phi\cdot\nabla\psi = -u_x u_y + u_y u_x = 0$:
equipotential lines and streamlines cross at right angles everywhere
the velocity is non-zero (Figure~\ref{fig:vol03ch11:potential-streamfunction}).
Third, the pairing is exactly the one that lets the two fields combine
into a single analytic complex potential $F(z) = \phi + i\psi$, the
construction the conformal-mapping box develops.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-potential-streamfunction.tex}

The orthogonality is not a curiosity; it is the basis of the
hand-drawn \engterm{flow net}, the pre-computer method for solving
Laplace's equation in a complicated geometry. The analyst sketches
streamlines and equipotentials so that they cross at right angles and
enclose curvilinear squares of roughly equal aspect ratio; where the
squares shrink the flow is fast, where they open out it is slow, and
the velocity is read straight off the spacing as $u \approx
\Delta\psi/\Delta n$ with $\Delta n$ the local square size. A flow net
drawn carefully gives velocities to a few percent, and it remains the
quickest way to develop intuition for a new geometry before a panel
code is written.

The linearity of Laplace's equation allows complex flows to be
constructed by superposing elementary solutions. Each elementary
solution corresponds to a basic flow type; their superposition
gives the flow around bodies of practical interest.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-potential-superposition.tex}

Figure~\ref{fig:vol03ch11:potential-superposition} collects the four
building blocks the rest of this section uses. The cylinder of the
last panel is a uniform stream plus a doublet; adding a vortex to it
is the entire content of the lifting-cylinder result that follows.

\engterm{Uniform flow}. A constant velocity $U_\infty$ in the
$x$-direction has potential and stream function
\[
\phi = U_\infty x, \quad \psi = U_\infty y.
\]
The streamlines are horizontal lines; the flow has no rotation, no
sources, no sinks.

\engterm{Source / sink}. A two-dimensional source of strength $m$
at the origin has
\[
\phi = \frac{m}{2\pi}\ln r, \quad \psi = \frac{m}{2\pi}\theta,
\]
in cylindrical coordinates $(r, \theta)$. The radial velocity is
$u_r = m/(2\pi r)$, decaying with distance from the source. A
sink is a negative source. Sources and sinks are mathematical
idealisations; real fluid injectors and drains have finite size
and a complicated near-field, but the far-field is well-described
by a source/sink solution.

\engterm{Doublet}. The limit of a source and sink approaching each
other with their strengths increasing such that the product
$\kappa = m d$ remains finite. The potential is
\[
\phi = -\frac{\kappa}{2\pi}\frac{\cos\theta}{r}, \quad
\psi = \frac{\kappa}{2\pi}\frac{\sin\theta}{r}.
\]
A doublet is the analogue of an electric dipole in electrostatics.

\engterm{Point vortex}. A two-dimensional point vortex of
circulation $\Gamma$ at the origin has
\[
\phi = \frac{\Gamma}{2\pi}\theta, \quad
\psi = -\frac{\Gamma}{2\pi}\ln r,
\]
giving a purely tangential velocity $u_\theta = \Gamma/(2\pi r)$.
The flow circulates around the vortex with a speed that decays
with distance; the vorticity is concentrated at the origin (a
delta function) and zero everywhere else. The point vortex is the
two-dimensional analogue of a smoke ring's core; real vortex cores
have finite size and a more complicated radial structure.

\subsection{The cylinder in uniform flow}

Superposing a uniform flow $U_\infty$ with a doublet of strength
$\kappa = 2\pi U_\infty R^2$ gives a flow around a circular
cylinder of radius $R$. The stream function is
\[
\psi = U_\infty r \sin\theta\left(1 - \frac{R^2}{r^2}\right),
\]
which vanishes on $r = R$ (the cylinder surface is a streamline)
and approaches $\psi = U_\infty y$ at large $r$ (uniform flow at
infinity). The velocity field is
\[
u_r = U_\infty \cos\theta\left(1 - \frac{R^2}{r^2}\right), \quad
u_\theta = -U_\infty \sin\theta\left(1 + \frac{R^2}{r^2}\right).
\]
On the cylinder surface ($r = R$), $u_r = 0$ and
$u_\theta = -2 U_\infty \sin\theta$. The flow speed on the surface
is twice the freestream at the top and bottom ($\theta = \pm\pi/2$)
and zero at the front and back stagnation points ($\theta = 0,
\pi$).

By Bernoulli, the pressure coefficient
$C_p = (p - p_\infty)/(\tfrac{1}{2}\rho U_\infty^2)$ on the
surface is
\[
C_p = 1 - 4\sin^2\theta.
\]
The pressure is maximum at the stagnation points ($C_p = 1$) and
minimum at the top and bottom ($C_p = -3$). Integrating the
pressure around the cylinder gives zero net force, in either the
streamwise or transverse direction: \engterm{d'Alembert's paradox}.
A cylinder in inviscid steady flow experiences no drag and no
lift, a result Jean le Rond d'Alembert noted in 1752 and that
foreshadowed the importance of viscosity (chapter 12 of this
volume) and unsteady wake structure (chapters 12 and 13) in real
fluid drag.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-cylinder-cp.tex}

Figure~\ref{fig:vol03ch11:cylinder-cp} plots $C_p = 1 - 4\sin^2\theta$
around the surface. The front-to-back symmetry of the curve is the
whole of the paradox: whatever the front pressure pushes against, the
rear pressure pushes back equally, and the streamwise integral
cancels. The script \path{code/cylinder_potential.py} integrates the
surface pressure numerically and recovers a drag indistinguishable
from zero (about $10^{-15}\,\si{\newton\per\meter}$, machine round-off)
for the non-lifting case, and a lift equal to $\rho U_\infty\Gamma$ to
the same precision once a circulation is added; the surface
distribution it writes to \path{data/cylinder_cp.csv} reproduces the
analytical curve. The paradox is not a defect of the arithmetic; it
is a true property of the inviscid model, and its resolution waits for
the boundary layer of chapter 12.

\begin{mastery}
\textbf{The complex potential and conformal mapping.} For
two-dimensional irrotational incompressible flow the velocity
potential $\phi$ and stream function $\psi$ are harmonic conjugates,
so they combine into a single analytic function of the complex
coordinate $z = x + iy$, the \engterm{complex potential}
\[
F(z) = \phi(x,y) + i\,\psi(x,y).
\]
Because $F$ is analytic, its derivative is the \engterm{complex
velocity}
\[
\frac{dF}{dz} = u_x - i\,u_y,
\]
a single differentiation that returns both velocity components. The
elementary flows take compact closed forms: a uniform stream is
$F = U_\infty z$, a source of strength $m$ is $F = (m/2\pi)\ln z$, a
vortex of circulation $\Gamma$ is $F = -i(\Gamma/2\pi)\ln z$, and a
doublet is $F = -\kappa/(2\pi z)$. The lifting cylinder is their sum,
\[
F(z) = U_\infty\!\left(z + \frac{R^2}{z}\right)
- \frac{i\Gamma}{2\pi}\ln z .
\]
The power of the complex form is \engterm{conformal mapping}: an
analytic map $\zeta = g(z)$ deforms one flow domain into another while
preserving Laplace's equation, so a flow solved in a simple geometry
transfers to a complicated one for the price of an algebraic
substitution. The \engterm{Joukowski transform}
$\zeta = z + c^2/z$ carries the circle of the lifting cylinder into a
family of airfoil-shaped contours, turning the exactly solvable
cylinder-with-circulation into an exactly solvable cambered airfoil,
the first analytic airfoil theory (Joukowski, 1910). The same
machinery solves flow around corners, through slots, and over steps;
it is the analytic counterpart of the panel methods the simulation
exercises build numerically.
\end{mastery}

\subsection{The cylinder with circulation}

Adding a point vortex of circulation $\Gamma$ at the centre of the
cylinder produces an asymmetric flow with a net transverse force.
The stream function becomes
\[
\psi = U_\infty r \sin\theta\left(1 - \frac{R^2}{r^2}\right)
- \frac{\Gamma}{2\pi}\ln\frac{r}{R}.
\]
The surface pressure distribution is no longer symmetric about
$\theta = 0$ to $\pi$: integration gives a transverse force per
unit length
\[
L' = \rho U_\infty \Gamma,
\]
the \engterm{Kutta-Joukowski theorem} (Martin Kutta, 1902; Nikolai
Joukowski, 1906). The lift per unit length of any two-dimensional
body in an inviscid uniform flow equals the product of the fluid
density, the freestream speed, and the bound circulation around the
body. The theorem is the analytical foundation of airfoil theory
and the working basis of subsonic aerodynamic design.

The result is worth deriving on the cylinder, where every step is
elementary and the general theorem is visible in the one term that
survives. With the vortex added, the surface speed becomes
$u_\theta = -2U_\infty\sin\theta - \Gamma/(2\pi R)$, the sum of the
non-lifting distribution and the uniform swirl the vortex contributes
at the radius $R$. Bernoulli gives the surface pressure
\[
p(\theta) = p_\infty + \tfrac12\rho U_\infty^2
- \tfrac12\rho\,u_\theta^2,
\]
and the transverse force per unit span is the integral of the
$y$-component of the pressure force, $-p\sin\theta$, around the
circumference,
\[
L' = -\int_0^{2\pi} p(\theta)\,\sin\theta\;R\,d\theta.
\]
Expanding $u_\theta^2 = 4U_\infty^2\sin^2\theta
+ 2U_\infty\sin\theta\,\Gamma/(\pi R) + \Gamma^2/(2\pi R)^2$, three
terms enter the integrand. The constant terms ($p_\infty$,
$\tfrac12\rho U_\infty^2$, and the $\Gamma^2$ term) integrate against
$\sin\theta$ to zero over a full turn. The $\sin^2\theta$ term
integrates against $\sin\theta$ to zero as well, by the odd symmetry of
$\sin^3\theta$. Only the cross term between the freestream and the
circulation, $-\tfrac12\rho\cdot 2U_\infty\sin\theta\cdot
\Gamma/(\pi R)$, survives, and it carries a $\sin^2\theta$ once the
explicit $\sin\theta$ of the projection is included:
\[
L' = -\int_0^{2\pi}\left[-\tfrac12\rho\,\frac{2U_\infty\Gamma}{\pi R}
\sin\theta\right]\sin\theta\;R\,d\theta
= \frac{\rho U_\infty\Gamma}{\pi}\int_0^{2\pi}\sin^2\theta\,d\theta
= \frac{\rho U_\infty\Gamma}{\pi}\cdot\pi
= \rho U_\infty\Gamma.
\]
The radius cancels, the freestream-only and circulation-only parts
contribute nothing, and the lift is the product $\rho U_\infty\Gamma$.
The same streamwise integration with $-p\cos\theta$ gives zero, the
d'Alembert result, because no term in the integrand survives against
$\cos\theta$. The lift lives entirely in the interference between the
oncoming stream and the bound vortex; this is why a body must carry
circulation to lift, and why the lift is exactly proportional to it.

\begin{mastery}
\textbf{The method of images: walls and free surfaces for free.} A
boundary that the flow cannot cross is a streamline, and the
linearity of Laplace's equation lets us build that streamline by
placing a mirror singularity outside the flow domain rather than
solving a boundary-value problem. The technique is borrowed intact
from electrostatics, where a grounded conducting plane is reproduced
by an image charge. Consider a point source of strength $m$ at height
$d$ above a flat solid wall along $y = 0$. The wall must carry zero
normal velocity. Placing an identical image source of strength $m$ at
$y = -d$, the mirror position, makes the two sources' vertical
velocities cancel exactly on $y = 0$ by symmetry, so the wall plane
is a streamline and the flow in the upper half-plane is the
superposition of the real source and its image. A vortex near a wall
takes an image vortex of opposite sign (so the tangential velocities
add and the normal velocities cancel), and the real vortex then
drifts parallel to the wall under the velocity its image induces, the
mechanism by which a wingtip vortex tracks along the ground during
takeoff and landing and by which a vortex ring approaching a wall
slows and spreads. A free surface, where the pressure rather than the
normal velocity is fixed, takes an image of the opposite kind: a
source images to a sink, so that the combined flow leaves the surface
undisturbed to first order. The same construction with a sequence of
images handles a source between two parallel walls (an infinite
image train) and a vortex inside a corner (a finite set of images
whose count depends on the corner angle). The method of images turns
a wall or a free surface into an algebra problem and is the first
tool an inviscid analyst reaches for whenever a boundary is flat.
\end{mastery}

\begin{archetype}[Balance, transport: Bernoulli plus Kutta-Joukowski]
The inviscid flow archetype puts two relations to work together.
Continuity sets the relationship between cross-sectional area and
flow speed; Bernoulli converts the speed variation into a pressure
variation; integrating the pressure over a body's surface gives the
net force. For a closed body in steady irrotational flow the
streamwise integration yields zero (d'Alembert's paradox); the
transverse integration yields the Kutta-Joukowski lift if the body
carries bound circulation. The archetype is the working basis of
every subsonic aerodynamics calculation, and it underwrites the
estimation habit of computing lift from a circulation estimate
rather than from an integration of surface pressures.
\end{archetype}

\begin{estimation}
\textbf{Estimate the pressure difference between the upper and
lower surfaces of an airliner wing at cruise, and compare it to
atmospheric pressure.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A working airliner cruises at altitude near $10\,\si{\kilo\meter}$
where the ambient pressure is approximately $26\,\text{kPa}$ and
the air density is about $0.41\,\si{\kilogram\per\meter\cubed}$.
The cruise speed is about $250\,\si{\meter\per\second}$ (Mach
$0.85$). A medium twin-aisle airliner has a maximum takeoff weight
of about $230\,\text{t}$ ($2.3\times 10^6\,\si{\newton}$ of weight)
and a wing area of about $360\,\si{\meter}^2$.

The wing must supply lift equal to the aircraft's weight in cruise:
$L = 2.3\times 10^6\,\si{\newton}$. The average lift per unit area
is then
\[
\frac{L}{A_w} = \frac{2.3\times 10^6}{360} \approx 6.4\,\text{kPa}.
\]
The pressure differential between upper and lower surfaces,
averaged over the wing area, is about $6.4\,\text{kPa}$, roughly
a quarter of the ambient pressure at altitude.

The dynamic pressure at cruise is
\[
q = \tfrac{1}{2}\rho U_\infty^2 \approx \tfrac{1}{2}\times 0.41
\times 250^2 \approx 12.8\,\text{kPa}.
\]
The average pressure-coefficient differential between upper and
lower surfaces is then $\Delta C_p \approx 6.4/12.8 = 0.5$. The
working airfoil lift coefficient at cruise is typically
$C_L \approx 0.4$ to $0.6$, consistent with this estimate (the
chord-averaged $\Delta C_p$ over the wing area equals $C_L$ by
definition for a thin airfoil).

The postmortem: the rough match between $\Delta C_p \approx 0.5$
and the textbook $C_L$ at cruise shows that the simple force
balance ``lift equals weight'' plus the Bernoulli definition of
$C_p$ gives a quantitatively reasonable picture of cruise flight.
The pressure differential is not small relative to ambient: a
quarter of an atmosphere across a wing's chord is the working
mechanical reality. The next chapter sets a viscous correction
that costs a few percent of cruise thrust; this chapter's inviscid
analysis recovers the lift to leading order.
\end{estimation}

\section{Worked examples}\pathtag{standard}

The worked problems of this section carry the bookkeeping in full. The
first group applies Bernoulli plus continuity to the standard
configurations of aerodynamics and internal flow, and to the
free-surface efflux cases (the tank-jet trajectory and the submerged
orifice) that turn Torricelli into a range and a two-reservoir
discharge; the next group adds the momentum theorem (jet impingement,
the moving Pelton bucket, the nozzle reaction force, the sudden
expansion) and the rotating-frame pressure balance (the forced vortex);
a third group works the superposition machinery harder (the Rankine
half-body and the closed Rankine oval, the thin-airfoil camber line as a
vortex sheet, the trailing-vortex pair, the unsteady oscillating column,
and the free-streamline contraction of a jet); a fourth set carries the
free-surface and suction-limit cases (the sluice gate and the siphon)
and a short angular-momentum vignette on the lawn sprinkler; the section
closes with three end-to-end studies, a Venturi metering station, a
water-hammer penstock, and a finite-wing induced-drag analysis, in which
the relations of the chapter are strung together into a sized result. A
reader who follows every step here can apply the inviscid toolkit to any
configuration that meets the five Bernoulli assumptions, and can
recognise from the momentum-theorem examples how the analysis continues
when one of them fails.

\subsection{Flow over an inviscid airfoil}

A symmetric thin airfoil of chord $c$ is placed at angle of attack
$\alpha$ in a uniform flow $U_\infty$. Thin-airfoil theory (Hermann
Glauert, 1926) approximates the airfoil as a line of bound vortices
distributed along the chord with circulation density $\gamma(x)$,
satisfying the boundary condition that the flow is tangent to the
camber line at every chord position. For a flat-plate symmetric
airfoil at small angle of attack, the result is
\[
C_L = 2\pi\alpha,
\]
the \engterm{thin-airfoil lift slope}. The lift coefficient is
linear in angle of attack with a slope of $2\pi$ per radian, or
about $0.11$ per degree.

The corresponding bound circulation per unit span is
\[
\Gamma = \pi c U_\infty \alpha,
\]
giving lift per unit span
\[
L' = \rho U_\infty \Gamma = \pi \rho U_\infty^2 c \alpha,
\]
in agreement with $L' = C_L \cdot \tfrac{1}{2}\rho U_\infty^2 c$
when $C_L = 2\pi\alpha$. The thin-airfoil result is the first
quantitative aerodynamic-design tool; real airfoils with finite
thickness and camber depart from it by perhaps $10$ to $20\,\%$ in
lift slope, more in the post-stall regime where the inviscid model
fails entirely (the topic of chapter 12 of this volume).

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-airfoil-circulation.tex}

Figure~\ref{fig:vol03ch11:airfoil-circulation} draws the
circulation picture the thin-airfoil result encodes: the freestream
and the bound circulation sum to a fast upper flow and a slow lower
flow, the pressure difference follows by Bernoulli, and the integrated
difference is the lift $\rho U_\infty\Gamma$. The Kutta condition is
the missing ingredient that fixes $\Gamma$; without it, the inviscid
equations admit any circulation and predict any lift.

\paragraph{Worked example: the lift slope as a sanity check.} A
flat-plate airfoil at $\alpha = 5^\circ = 0.0873\,\text{rad}$ in a
$60\,\si{\meter\per\second}$ stream of density
$1.2\,\si{\kilogram\per\meter\cubed}$ and chord $1\,\si{\meter}$ has
$C_L = 2\pi\alpha = 0.548$. The lift per unit span is $L' = C_L\cdot
\tfrac12\rho U_\infty^2 c = 0.548\times 0.5\times 1.2\times 60^2\times
1 = 1184\,\si{\newton\per\meter}$, and the bound circulation is
$\Gamma = L'/(\rho U_\infty) = 1184/(1.2\times 60) =
16.4\,\si{\meter\squared\per\second}$. The same $\Gamma$ follows
directly from $\Gamma = \pi c U_\infty\alpha = \pi\times 1\times
60\times 0.0873 = 16.4\,\si{\meter\squared\per\second}$, confirming
that the lift-slope and circulation routes agree. The two independent
routes meeting is the check; a calculation in which they disagree has
an arithmetic error or a hidden inconsistency in the assumptions.

\subsection{Nozzle flow}

A converging nozzle reduces the cross-sectional area of a steady
incompressible flow from $A_1$ to $A_2$, accelerating the flow from
speed $u_1$ to $u_2 = u_1(A_1/A_2)$ and decreasing the pressure
from $p_1$ to $p_2 = p_1 - \tfrac{1}{2}\rho(u_2^2 - u_1^2)$.

For a converging-diverging nozzle the flow accelerates in the
converging section, reaches its maximum speed at the throat (the
minimum area), and decelerates in the diverging section, recovering
some of the pressure. In an inviscid analysis the pressure recovery
is complete: the exit pressure equals the inlet pressure if the
exit area equals the inlet area. The contraction and the expansion are
not symmetric in their difficulty, and the asymmetry is the whole of
why a nozzle is easy and a diffuser is hard. In the converging section
the pressure falls in the flow direction, a \engterm{favourable
pressure gradient}: the fluid is pushed from behind toward a lower
pressure ahead, the flow accelerates, and the thin wall layer stays
firmly attached, so an inviscid contraction and a real one behave
almost identically and a nozzle can turn a right angle of area change
in a short length with negligible loss. In the diverging section the
pressure rises in the flow direction, an \engterm{adverse pressure
gradient}: the near-wall fluid, already slowed by whatever friction it
has met, must climb toward a higher pressure ahead, and if the rise is
steep the slow fluid runs out of momentum, stalls, and reverses, so the
flow separates from the wall and the pressure recovery collapses into a
turbulent, dissipative wake. This is why the pressure recovery a real
diffuser achieves is quoted as a \engterm{pressure-recovery
coefficient} $C_{pr} = (p_\text{exit} - p_\text{throat})/(\tfrac12\rho
u_\text{throat}^2)$ that a designer works to raise toward the inviscid
ideal $1 - (A_\text{throat}/A_\text{exit})^2$ but never reaches; a
well-designed diffuser recovers about $70$ to $90\,\%$ of the dynamic
pressure expended in the contraction, and a badly proportioned one,
opened too fast, recovers far less because it separates. The discipline
of diffuser design is to keep the divergence angle below the separation
threshold (about $7^{\circ}$ on each side for unaided flow, more with
vortex generators or boundary-layer trips), and to accept that a
diffuser buying a given area ratio must be several times longer than the
contraction that produced the speed, because the adverse gradient must
be spread over enough length to keep the wall layer attached. The
sudden-expansion example later in this section is the limit of a
diffuser opened instantly, where the entire recoverable pressure is
thrown away in the separated corner.

For a compressible flow the analysis changes qualitatively at the
throat: above a critical pressure ratio, the throat \engterm{
chokes} and the flow speed at the throat equals the local speed of
sound. Beyond the throat, a converging-diverging nozzle can
accelerate the flow supersonically. Chapter 13 of this volume
develops the compressible analysis; the incompressible Bernoulli
result of this chapter is accurate at Mach numbers below about
$0.3$.

\subsection{Dam spillway}

Water flows over the crest of a dam spillway and accelerates down
the spillway face. Applying Bernoulli from the reservoir (where the
water is nearly at rest at elevation $z_1$) to a point on the
spillway face at elevation $z_2$, with both points at atmospheric
pressure,
\[
\tfrac{1}{2} u_2^2 = g(z_1 - z_2),
\]
giving $u_2 = \sqrt{2g(z_1 - z_2)}$. The spillway accelerates the
water through a head equal to the difference in elevation; this is
Torricelli's law applied to a free-surface flow.

The discharge per unit width of crest is set by the crest geometry
and the head above the crest. For a sharp-crested rectangular weir
of width $b$ and head $H$ above the crest,
\[
Q = C_d \tfrac{2}{3} b\sqrt{2g}\,H^{3/2},
\]
where $C_d \approx 0.62$ is a discharge coefficient that accounts
for contraction of the jet at the crest. The $H^{3/2}$ power is not a
fitted exponent; it falls out of Bernoulli integrated over the depth of
the overfalling sheet, and the integration is short enough to carry.
Treat the crest as an opening whose top is the free surface and whose
bottom is the crest, and let $\eta$ measure depth below the free
surface, so $\eta$ runs from $0$ at the surface to $H$ at the crest. A
fluid element at depth $\eta$ has fallen a head $\eta$ from the nearly
still upstream pool, so by Bernoulli it crosses the crest at speed
$\sqrt{2g\eta}$. The discharge is that speed integrated over the sheet,
each horizontal strip of width $b$ and thickness $d\eta$ carrying
$b\sqrt{2g\eta}\,d\eta$,
\[
Q_\text{ideal} = \int_0^H b\sqrt{2g\eta}\,d\eta
= b\sqrt{2g}\,\Big[\tfrac{2}{3}\eta^{3/2}\Big]_0^H
= \tfrac{2}{3}b\sqrt{2g}\,H^{3/2},
\]
and the discharge coefficient $C_d$ multiplies this ideal to absorb the
jet contraction at the crest and the small viscous loss. The three-
halves power is the product of the velocity's half power of head and the
area's first power of head: the flow runs faster and through a taller
gap as the head rises, and the two effects multiply. The steepness of
that power is what makes a weir a sensitive gauge, a ten percent rise in
head lifting the discharge by about fifteen percent, and also what makes
a spillway a self-relieving safety device, a small rise in reservoir
level passing a large increase in flood flow.

The same free-surface Bernoulli underlies a second reading of the
spillway, the one that explains why the crest, and not some point
upstream or down the face, is where the depth is set. In an open channel
the sum of the depth and the velocity head, the \engterm{specific
energy} $E = y + u^2/(2g)$ measured from the bed, has a minimum at a
particular depth for a given discharge, the \engterm{critical depth}
$y_c = (q^2/g)^{1/3}$ with $q$ the flow per unit width. Flow deeper and
slower than critical is \engterm{subcritical} and is controlled from
downstream; flow shallower and faster is \engterm{supercritical} and is
controlled from upstream. A spillway crest is the point where the flow
passes through critical, so the crest fixes the depth and the head-
discharge relation is set there regardless of conditions further down
the face, where the accelerating supercritical sheet is governed by the
$u = \sqrt{2g(z_1 - z_2)}$ free-fall balance already written. The
hydraulic jump of the failure section is the return passage from
supercritical back to subcritical at the toe, and it is the one place on
the whole spillway where Bernoulli fails and the momentum balance must
take over. Dam-engineering practice uses extensive empirical tables of
$C_d$ for various spillway geometries (ogee crests, broad-crested weirs,
side channels, chute spillways); the underlying physics is Bernoulli
plus continuity through the critical section, with empirical corrections
for viscosity, surface roughness, and separation effects. A spillway
designed for a $1000$-year flood will be sized using these relations;
the design uncertainty is mainly in the flood event, not in the
Bernoulli analysis.

\subsection{Draining-tank time as an unsteady Bernoulli problem}

Torricelli gives the instantaneous efflux speed; the time to drain a
tank turns that instantaneous law into an ordinary differential
equation, and the integration is worth carrying through because the
square-root-of-head law produces a drain time that surprises the
intuition built on constant-rate emptying.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-draining-tank.tex}

\paragraph{Worked example: draining a swimming pool.} A rectangular
pool of plan area $A_s = 25\times 10 = 250\,\si{\meter}^2$, filled to
depth $h_0 = 2\,\si{\meter}$, drains through a bottom orifice of
diameter $100\,\si{\milli\meter}$ ($A_o = \tfrac{\pi}{4}(0.1)^2 =
7.85\times 10^{-3}\,\si{\meter}^2$) with discharge coefficient $C_d =
0.62$ (Figure~\ref{fig:vol03ch11:draining-tank}). Volume conservation
equates the rate the surface drops to the rate volume leaves the
orifice:
\[
A_s\,\frac{dh}{dt} = -C_d A_o \sqrt{2 g h}.
\]
This is separable. Writing $\beta = C_d A_o \sqrt{2g}/A_s$,
\[
\frac{dh}{\sqrt{h}} = -\beta\,dt
\quad\Longrightarrow\quad
2\sqrt{h} = 2\sqrt{h_0} - \beta t,
\]
so the depth history is the parabola $h(t) = \big(\sqrt{h_0} -
\tfrac{\beta}{2}t\big)^2$, and the tank is empty ($h = 0$) at
\[
T = \frac{2\sqrt{h_0}}{\beta}
= \frac{A_s}{C_d A_o}\sqrt{\frac{2 h_0}{g}}.
\]
Putting in numbers, $\sqrt{2 h_0/g} = \sqrt{4/9.81} =
0.639\,\si{\second}$ (carrying $\si{\meter}^{1/2}$ implicitly), and
$A_s/(C_d A_o) = 250/(0.62\times 7.85\times 10^{-3}) = 5.14\times
10^4$, so
\[
T = 5.14\times 10^4 \times 0.639 = 3.28\times 10^4\,\si{\second}
\approx 9.1\,\text{h}.
\]
A pool that would empty in about three hours at the initial efflux
rate of $C_d A_o\sqrt{2 g h_0} = 0.62\times 7.85\times 10^{-3}\times
6.26 = 0.030\,\si{\meter\cubed\per\second}$ (which clears the
$500\,\si{\meter}^3$ in $1.6\times 10^4\,\si{\second}\approx
4.6\,\text{h}$) in fact takes twice as long, because the head, and
with it the efflux speed, falls as the tank empties. The script
\path{code/tank_drain.py} integrates the same ODE numerically and
recovers the nine-hour figure, a check that the closed form and the
numerics agree.

\subsection{The trajectory and range of a tank jet}

Torricelli fixes the speed of a jet leaving a tank; where the jet lands
is then a projectile question, and pairing the two answers the everyday
puzzle of which hole in a leaking tank throws its water farthest. The
example is worth carrying because the answer is not the deepest hole,
and the reason is a clean product of the efflux law and free fall.

\paragraph{Worked example: the farthest-reaching leak.} A tank stands
with its water surface a height $H = 1.8\,\si{\meter}$ above the ground,
and a small hole is punched in the vertical side wall at a depth $h$
below the surface, so the hole is a height $H - h$ above the ground. The
jet leaves horizontally at the Torricelli speed $u = \sqrt{2gh}$ and
then falls freely. The time to reach the ground from height $H - h$ is
$t = \sqrt{2(H - h)/g}$, and the horizontal range is
\[
x = u\,t = \sqrt{2gh}\,\sqrt{\frac{2(H - h)}{g}}
= 2\sqrt{h(H - h)}.
\]
The range is largest where the product $h(H - h)$ is largest, which by
symmetry is $h = H/2$, the hole at mid-depth. For $H = 1.8\,\si{\meter}$
the best hole sits at $h = 0.9\,\si{\meter}$ and throws the jet a
horizontal distance $x = 2\sqrt{0.9\times 0.9} = 2\times 0.9 =
1.8\,\si{\meter}$, exactly the surface height. A hole near the bottom
($h$ large) leaves at high speed but has almost no fall time; a hole
near the surface ($h$ small) has plenty of fall time but almost no
speed; the mid-depth hole balances the two. Two holes placed
symmetrically about mid-depth, at $h$ and $H - h$, throw their jets to
the same range, because $h(H - h)$ is unchanged by the swap. The result
is a Torricelli-plus-kinematics pairing with no fitting constant, and it
is the standard demonstration that efflux speed alone does not decide
where a leak lands.

\subsection{A submerged orifice between two reservoirs}

The Torricelli law took the jet discharging to the atmosphere, so the
driving head was the depth below the free surface. When an orifice
connects two bodies of liquid, the jet discharges into liquid rather
than into air, and the driving head is the difference of the two free
surfaces, not the depth below either. The distinction is worth one
example because a designer who uses the upstream depth in place of the
level difference oversizes the flow badly.

\paragraph{Worked example: flow through a submerged sluice orifice.} A
bulkhead separates two large basins whose free surfaces stand at
different levels, the upstream surface a height $\Delta H =
1.2\,\si{\meter}$ above the downstream surface. A submerged orifice of
area $A_o = 0.05\,\si{\meter}^2$ and discharge coefficient $C_d = 0.61$
pierces the bulkhead well below both surfaces. Apply Bernoulli on a
streamline from a point on the upstream surface (at rest, atmospheric)
to a point in the jet just downstream of the orifice. The downstream
jet is surrounded by the downstream water, so its pressure is the local
hydrostatic pressure at the orifice depth, set by the downstream
surface; the static pressures at the two endpoints therefore differ by
the hydrostatic head of the level difference, and the elevation and
pressure terms combine so that the net driving head is simply $\Delta
H$. The jet speed is
\[
u = \sqrt{2 g\,\Delta H}
= \sqrt{2\times 9.81\times 1.2}
= 4.85\,\si{\meter\per\second},
\]
and the discharge is $Q = C_d A_o u = 0.61\times 0.05\times 4.85 =
0.148\,\si{\meter\cubed\per\second}$. The depth of the orifice below
either surface never enters, only the difference of levels. Had the
designer instead used the upstream depth below its own surface, say
$3\,\si{\meter}$, the speed would come out $\sqrt{2\times 9.81\times 3}
= 7.7\,\si{\meter\per\second}$ and the flow about sixty percent high,
because the downstream water pushes back with its own hydrostatic
pressure and cancels all but the level difference. A submerged orifice
meters the head between two surfaces; a free-discharging orifice meters
the depth below one.

\subsection{The moving Pelton bucket and the half-speed power optimum}

The jet-vane example held the vane fixed and found the force. A working
turbine bucket moves, and the power it collects depends on the bucket
speed through the relative velocity the bucket sees. Carrying the moving
case shows why a Pelton wheel is run at half the jet speed, the single
design rule that sets the runner diameter for a given head.

\paragraph{Worked example: the runner speed for maximum power.} A free
jet of speed $u$ and mass flow $\dot m$ (fixed by the nozzle, since the
jet is set before it reaches the wheel) strikes a bucket moving away
from the nozzle at speed $v$. In the bucket's frame the jet arrives at
the relative speed $u - v$, is turned through nearly $180^\circ$, and
leaves at the same relative speed $u - v$ for an inviscid bucket. The
mass flux the moving bucket intercepts is the full nozzle flow $\dot m$
for a wheel with many buckets (each catching the jet in turn), so the
force on the bucket is $F = \dot m\,(1 - \cos\theta)(u - v)$ with
$\theta$ the turning angle, and the power delivered is the force times
the bucket speed,
\[
P = F\,v = \dot m\,(1 - \cos\theta)\,(u - v)\,v.
\]
The power is a downward parabola in $v$: zero at $v = 0$ (a stationary
bucket delivers force but no power) and zero again at $v = u$ (a bucket
running with the jet feels no relative flow and collects nothing).
Between them the maximum is where $dP/dv = 0$,
\[
\frac{dP}{dv} = \dot m\,(1 - \cos\theta)\,(u - 2v) = 0
\quad\Longrightarrow\quad
v = \frac{u}{2}.
\]
The bucket delivers most power when it runs at half the jet speed. At
that speed the power is $P_\text{max} = \dot m\,(1 - \cos\theta)\,
(u/2)(u/2) = \tfrac14\dot m\,(1 - \cos\theta)\,u^2$, and with a
near-full reversal $\theta = 165^\circ$, $(1 - \cos\theta) = 1.97$, so
$P_\text{max} = 0.49\,\dot m\,u^2$, close to the whole kinetic-energy
flux $\tfrac12\dot m u^2$ of the jet. The inviscid moving bucket
converts almost all of the jet's kinetic energy to shaft power at the
half-speed condition, which is why a Pelton installation picks the
runner diameter and rotational speed so that the bucket circumferential
speed lands near half the spouting velocity $\sqrt{2gH}$ of its head.

\subsection{Pitot-static airspeed and the compressibility correction}

The incompressible Pitot relation $u = \sqrt{2(p_0 - p)/\rho}$ is the
working tool of low-speed anemometry; at flight Mach numbers the same
measurement needs a correction, and seeing where the correction
enters is the point of the example.

\paragraph{Worked example: airspeed at low and at high subsonic
Mach.} A Pitot-static probe reads a stagnation-static differential of
$\Delta p = p_0 - p = 4.0\,\text{kPa}$ in air of static pressure $p =
70\,\text{kPa}$ and density $\rho = 0.91\,\si{\kilogram\per\meter
\cubed}$ (about $3\,\si{\kilo\meter}$ altitude). The incompressible
reading is
\[
u = \sqrt{\frac{2\Delta p}{\rho}} = \sqrt{\frac{2\times 4000}{0.91}}
= \sqrt{8791} = 93.8\,\si{\meter\per\second}.
\]
The local speed of sound is $c = \sqrt{\gamma p/\rho} = \sqrt{1.4
\times 70\,000/0.91} = \sqrt{1.077\times 10^5} =
328\,\si{\meter\per\second}$, so the Mach number is $M = u/c = 93.8/
328 = 0.29$, just at the edge of the incompressible regime. The
compressible stagnation relation for isentropic flow of a perfect gas
is
\[
\frac{p_0}{p} = \left(1 + \frac{\gamma - 1}{2}M^2\right)
^{\gamma/(\gamma - 1)},
\]
which at $M = 0.29$ gives $p_0/p = (1 + 0.2\times 0.0841)^{3.5} =
(1.0168)^{3.5} = 1.0601$, against the incompressible prediction
$1 + \tfrac12\gamma M^2 = 1 + 0.7\times 0.0841 = 1.0589$. The two
agree to about $0.1\,\%$ in pressure ratio at this Mach number, so
the incompressible airspeed is good to a fraction of a percent here.
Repeating at $M = 0.7$ (a transport cruise condition), the
compressible ratio is $(1 + 0.2\times 0.49)^{3.5} = (1.098)^{3.5} =
1.387$ while the incompressible form gives $1 + 0.7\times 0.49 =
1.343$, a $3\,\%$ discrepancy in pressure ratio that translates to a
few percent in computed speed. The lesson sets the boundary: below
$M \approx 0.3$ the incompressible Pitot relation is accurate to under
a percent; above it the compressible stagnation relation of
chapter 13 of this volume must replace it, and an airspeed indicator
that ignores the correction reads progressively low as Mach rises.

\subsection{Orifice metering and the cavitation margin}

A metering orifice sized for a clean differential-pressure signal can
inadvertently drive its own throat into cavitation; the example
closes the loop between the Venturi/orifice metering physics and the
cavitation failure mode by carrying both calculations on one device.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-orifice-vena-contracta.tex}

\paragraph{Worked example: an orifice plate near its cavitation
limit.} An orifice plate of bore diameter $40\,\si{\milli\meter}$
meters water in a $100\,\si{\milli\meter}$ line at $Q =
12\,\text{L/s}$, with the line at a static pressure of $p_1 =
180\,\text{kPa}$ absolute (Figure~\ref{fig:vol03ch11:orifice-vena-contracta}).
The bore area is $A_2 = \tfrac{\pi}{4}(0.04)^2 = 1.257\times
10^{-3}\,\si{\meter}^2$, and the flow contracts further to a vena
contracta of area $A_c \approx C_c A_2$ with contraction coefficient
$C_c \approx 0.62$, so $A_c = 7.79\times 10^{-4}\,\si{\meter}^2$. The
speed at the vena contracta is
\[
u_c = \frac{Q}{A_c} = \frac{0.012}{7.79\times 10^{-4}}
= 15.4\,\si{\meter\per\second}.
\]
The approach speed in the line is $u_1 = Q/A_1 = 0.012/(7.85\times
10^{-3}) = 1.53\,\si{\meter\per\second}$, negligible by comparison.
Bernoulli from the line to the vena contracta gives the minimum
pressure
\[
p_\text{min} = p_1 - \tfrac12\rho(u_c^2 - u_1^2)
= 180\,000 - 0.5\times 1000\times(237 - 2.3)
= 180\,000 - 117\,400 = 62.6\,\text{kPa}.
\]
The cavitation margin is the gap to the vapour pressure $p_v \approx
2.3\,\text{kPa}$ absolute, expressed as the cavitation number
\[
\sigma = \frac{p_\text{min} - p_v}{\tfrac12\rho u_c^2}
= \frac{62\,600 - 2300}{117\,400} = 0.51.
\]
For sharp-edged orifices the inception value of $\sigma$ is typically
near $0.6$ to $1.5$ depending on the recovery geometry, so $\sigma =
0.51$ sits inside the cavitating range: this orifice cavitates at this
flow, and the downstream face and pipe wall will pit over time. The
fix is to lower the velocity at the contraction by opening the bore
(a $50\,\si{\milli\meter}$ bore drops $u_c$ to about
$9.9\,\si{\meter\per\second}$ and $\tfrac12\rho u_c^2$ to
$49\,\text{kPa}$, lifting $\sigma$ above $2.5$), at the cost of a
smaller and noisier differential signal, or to raise the line
pressure with downstream backpressure. The metering accuracy and the
cavitation margin pull against each other, and the design must satisfy
both.

\begin{estimation}
\textbf{Estimate the velocity and the cavitation risk at the foot of
a $30\,\si{\meter}$ dam spillway.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
Water released at the crest is nearly at rest; at the foot it has
fallen a head of about $30\,\si{\meter}$. Torricelli on the
free-surface streamline gives $u = \sqrt{2gh} = \sqrt{2\times 9.81
\times 30} \approx 24\,\si{\meter\per\second}$. At that speed the
dynamic pressure is $q = \tfrac12\rho u^2 = 0.5\times 1000\times 24^2
\approx 290\,\text{kPa}$.

Where the high-speed sheet passes over a surface irregularity (a
construction joint, an offset, a slot for a gate), the local static
pressure can drop by a sizeable fraction of $q$ below the ambient
atmospheric value of about $100\,\text{kPa}$ absolute. A local
$C_p$ of only $-0.4$ at that speed gives a pressure drop of
$0.4\times 290 = 116\,\text{kPa}$, pulling the local absolute pressure
to roughly $100 - 116 < 0$, below the vapour pressure of about
$2.3\,\text{kPa}$: cavitation. The cavitation number there is
$\sigma = (100 - 2.3)/290 \approx 0.34$, near the inception range for
surface irregularities.

\smallskip
The postmortem: the estimate explains why high-head spillways are
aerated. Introducing air at the surface raises the local pressure and
cushions bubble collapse; the design discipline is to keep $\sigma$
above the inception value at every irregularity, or to aerate where
that cannot be guaranteed. The Bernoulli pressure drop is the
quantity that flags the hazard, and a few tens of metres of head is
enough to reach it.
\end{estimation}

\subsection{Jet impingement: the force a turned jet exerts on a vane}

Bernoulli fixes the speed of a free jet; the momentum theorem turns
that speed into a force when the jet strikes a surface. The pairing is
the working content of impulse turbines, thrust reversers, and the
recoil of a fire hose, and the bookkeeping is worth carrying once in
full because the sign of the turning angle decides the answer.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-jet-vane.tex}

\paragraph{Worked example: a water jet on a Pelton bucket.} A free
water jet of diameter $d = 30\,\si{\milli\meter}$ leaves a nozzle at
$u_1 = 40\,\si{\meter\per\second}$ and strikes a fixed vane that turns
it through $\theta = 165^\circ$, the working deflection of a Pelton
bucket (Figure~\ref{fig:vol03ch11:jet-vane}). The jet is at
atmospheric pressure along its whole path, so Bernoulli on a
free-surface streamline keeps the speed unchanged across an inviscid
vane: $u_2 = u_1 = 40\,\si{\meter\per\second}$. The jet area is $A =
\tfrac{\pi}{4}(0.03)^2 = 7.07\times 10^{-4}\,\si{\meter}^2$ and the
mass flow is $\dot m = \rho A u_1 = 1000\times 7.07\times 10^{-4}\times
40 = 28.3\,\si{\kilogram\per\second}$. Take $x$ along the incoming
jet. The outgoing momentum flux has $x$-component $\dot m\,u_2\cos
\theta$ (with $\cos 165^\circ = -0.966$, the jet is thrown back), so
the $x$-momentum balance gives the force the vane exerts on the fluid,
$F_x = \dot m(u_2\cos\theta - u_1) = 28.3\times 40\times(-0.966 - 1) =
-2226\,\si{\newton}$. The reaction on the vane is $+2226\,\si{\newton}$
in the direction of the jet, the force that drives the wheel. A flat
plate normal to the jet ($\theta = 90^\circ$, $\cos\theta = 0$) would
collect only $\dot m\,u_1 = 1132\,\si{\newton}$; the deep bucket nearly
doubles the force because it reverses the momentum rather than merely
stopping it. The factor $1 - \cos\theta$ is the whole of the design
case for a cupped bucket over a flat paddle, and it is bounded by
$2$ at full reversal. A moving bucket collects less, because the
mass-flow it intercepts and the relative turning speed both fall with
the runner speed; the maximum power transfers when the bucket runs at
half the jet speed, the classical Pelton result.

\subsection{The forced vortex: a spinning liquid's paraboloidal surface}

A liquid rotating with its container as a rigid body carries no shear
and no relative motion, so the inviscid pressure field describes it
exactly even though a real fluid reaches solid-body rotation only
through viscosity. The radial pressure gradient supplies the
centripetal acceleration, and the free surface, a constant-pressure
surface, takes the shape that the pressure field dictates.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-forced-vortex.tex}

\paragraph{Worked example: the surface of a spun bucket.} A cylindrical
bucket of radius $R = 0.25\,\si{\meter}$ holds water and is spun about
its axis at $\omega = 12\,\si{\radian\per\second}$. In the rotating
frame the radial momentum balance reads $\partial p/\partial r =
\rho\,\omega^2 r$ (the pressure gradient that turns the fluid),
while the vertical balance stays hydrostatic, $\partial p/\partial z =
-\rho g$. Integrating, $p(r,z) = p_0 + \tfrac12\rho\omega^2 r^2 -
\rho g z$. The free surface is the locus $p = p_\text{atm}$, which
rearranges to
\[
z(r) = z_0 + \frac{\omega^2 r^2}{2g},
\]
the paraboloid of Figure~\ref{fig:vol03ch11:forced-vortex}. The rise
from the axis to the wall is $\Delta z = \omega^2 R^2/(2g) = 12^2\times
0.25^2/(2\times 9.81) = 144\times 0.0625/19.62 = 0.459\,\si{\meter}$.
Volume conservation fixes the centre depression and the wall climb
symmetrically about the rest level: the paraboloid's mean height over
the area equals the still-water height, and for a paraboloid the axis
sits $\tfrac12\Delta z = 0.229\,\si{\meter}$ below the rest level while
the wall sits the same amount above it. A designer who spins an open
vessel must allow this much freeboard or the liquid climbs the wall and
spills; the same relation, read the other way, is how a spin-cast
furnace forms a parabolic mirror blank by rotating molten glass until
it sets, the rotation rate setting the focal length $f = g/(2\omega^2)$.
At $\omega = 12\,\si{\radian\per\second}$ that focal length is
$9.81/(2\times 144) = 0.034\,\si{\meter}$, a fast mirror; a long-focus
astronomical blank needs a slow spin.

\subsection{Sudden expansion: the Borda-Carnot loss the momentum theorem captures}

A pipe that widens abruptly is the one duct where the momentum theorem
and Bernoulli give different answers, and the difference is exactly the
energy the expansion dissipates. The case earns its place because it
shows the diagnostic of the failure section at work inside a worked
calculation: momentum is conserved across the expansion, mechanical
energy is not, and the gap between the two is a predictable loss.

\paragraph{Worked example: pressure rise across a sudden expansion.}
Water flows at $u_1 = 4\,\si{\meter\per\second}$ in a
$50\,\si{\milli\meter}$ pipe that opens abruptly into a
$100\,\si{\milli\meter}$ pipe. Continuity gives the downstream speed
from the area ratio $A_1/A_2 = (50/100)^2 = 0.25$, so $u_2 = 0.25\times
4 = 1.0\,\si{\meter\per\second}$. The jet of fast fluid leaving the
small pipe does not fill the large pipe at once; it expands through a
recirculating annulus, and over that region the pressure on the
back-facing step acts on the full large-pipe area. The momentum
theorem on a control volume spanning the expansion, taking the step
pressure equal to the upstream static pressure $p_1$ across the whole
face $A_2$ (the standard Borda assumption), gives
\[
p_1 A_2 - p_2 A_2 = \dot m(u_2 - u_1) = \rho A_2 u_2 (u_2 - u_1),
\]
so the pressure rise is $p_2 - p_1 = \rho u_2(u_1 - u_2) = 1000\times
1.0\times(4 - 1) = 3000\,\text{Pa}$. Bernoulli, were it valid, would
predict a larger rise, $p_2 - p_1 = \tfrac12\rho(u_1^2 - u_2^2) =
0.5\times 1000\times(16 - 1) = 7500\,\text{Pa}$. The shortfall is the
\engterm{Borda-Carnot head loss}
\[
\Delta p_\text{loss} = \tfrac12\rho(u_1 - u_2)^2 = 0.5\times 1000\times
(3)^2 = 4500\,\text{Pa},
\]
and the bookkeeping closes exactly: $7500 = 3000 + 4500$. The loss is
the kinetic energy of the velocity difference, dissipated in the
turbulence of the recirculating corner. The momentum theorem made no
inviscid assumption and gave the true pressure rise; Bernoulli assumed
a loss-free transition and overpredicted it by the dissipated amount.
A diffuser exists precisely to avoid this loss by expanding the flow
gently enough that it stays attached; the sudden expansion is the limit
in which all of the recoverable pressure is thrown away.

\subsection{The Rankine half-body from a source in a stream}

Superposing a point source on a uniform stream produces the simplest
closed-form model of a streamlined nose, and the derivation exercises
the superposition machinery on a body whose shape is not assumed in
advance but emerges from the flow. The example is worth carrying
through because the dividing streamline that bounds the body is found,
not posited, and the surface pressure that a Bernoulli calculation
then needs follows from a velocity field written in closed form.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-rankine-halfbody.tex}

\paragraph{Worked example: nose of a bridge pier facing a current.} A
two-dimensional source of strength $m$ sits at the origin in a uniform
stream $U_\infty$ along $x$ (Figure~\ref{fig:vol03ch11:rankine-halfbody}).
The combined stream function is the sum of the two elementary stream
functions,
\[
\psi = U_\infty r\sin\theta + \frac{m}{2\pi}\theta,
\]
and the velocity components follow by differentiation,
\[
u_r = U_\infty\cos\theta + \frac{m}{2\pi r}, \qquad
u_\theta = -U_\infty\sin\theta.
\]
The flow stagnates where both components vanish. The tangential
component is zero on the axis ($\theta = \pi$, the upstream
direction), and there the radial component vanishes at
\[
r = a = \frac{m}{2\pi U_\infty},
\]
the stagnation point a distance $a$ upstream of the source, where the
source's outward push exactly cancels the oncoming stream. The
streamline through the stagnation point carries $\psi =
U_\infty\cdot 0 + (m/2\pi)\pi = m/2$, and this dividing streamline,
\[
U_\infty r\sin\theta + \frac{m}{2\pi}\theta = \frac{m}{2},
\]
is the contour the freestream cannot cross: read as a solid surface,
it is the \engterm{Rankine half-body}. Solving for the contour radius,
$r = a(\pi - \theta)/\sin\theta$, gives a smooth nose at $\theta =
\pi$ that opens downstream to an asymptotic half-width $y \to \pi a$ as
$\theta \to 0$, because all the source flux $m$ must eventually flow
inside the contour and $m = U_\infty\cdot 2\pi a = U_\infty\cdot(2)(\pi
a)$ fixes the far-field width at $\pm\pi a$. Take a pier nose with $a =
0.5\,\si{\meter}$ in a river current of $U_\infty =
2\,\si{\meter\per\second}$, so $m = 2\pi U_\infty a = 2\pi\times 2
\times 0.5 = 6.28\,\si{\meter\squared\per\second}$. The maximum
surface speed occurs near the shoulder of the nose; evaluating the
speed $|\vec u|^2 = u_r^2 + u_\theta^2$ on the contour and maximising
gives a peak of about $1.26\,U_\infty = 2.52\,\si{\meter\per\second}$
near $\theta \approx 63^\circ$, so the largest surface suction by
Bernoulli is $p_\text{min} - p_\infty = -\tfrac12\rho(|\vec
u|_\text{max}^2 - U_\infty^2) = -0.5\times 1000\times(6.35 - 4.0) =
-1.18\,\text{kPa}$ below the stream static pressure. The pier nose
runs a modest suction, well clear of cavitation in a river, and the
stagnation pressure on the nose tip is the full $\tfrac12\rho
U_\infty^2 = 2\,\text{kPa}$ above ambient, the load a debris strike
would add to. The shape itself fell out of the superposition; no nose
profile was assumed.

\subsection{The Rankine oval: a closed body from a source-sink pair}

The half-body of the last example stayed open downstream because a
lone source has nowhere to return its flux. Adding an equal sink a
short distance behind the source recaptures every streamline the
source emits, and the dividing streamline closes into a finite body,
the \engterm{Rankine oval}. The example is worth carrying because it
gives, in closed form, a complete two-dimensional body whose length
and thickness a designer can tune with two numbers, and because the
same superposition collapses to the cylinder in one limit and stretches
to a slender strut in the other.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-rankine-oval.tex}

\paragraph{Worked example: a rounded strut in a water tunnel.} Place a
source of strength $m$ at $x = -a$ and a sink of strength $-m$ at $x =
+a$ in a uniform stream $U_\infty$ along $x$
(Figure~\ref{fig:vol03ch11:rankine-oval}). The combined stream
function is the stream plus the two elementary parts, each measured by
its own angle $\theta_1$, $\theta_2$ from the source and the sink,
\[
\psi = U_\infty y + \frac{m}{2\pi}(\theta_1 - \theta_2).
\]
On the $x$-axis outside the pair the source and sink contributions to
the axial velocity add, and the flow stagnates where the net axial
velocity vanishes,
\[
U_\infty + \frac{m}{2\pi}\!\left(\frac{1}{x + a} - \frac{1}{x - a}
\right) = 0
\quad\Longrightarrow\quad
U_\infty = \frac{m}{2\pi}\,\frac{2a}{x^2 - a^2},
\]
so the two stagnation points sit at $x = \pm L$ with $L = a\sqrt{1 +
m/(\pi U_\infty a)}$, the half-length of the oval. Take a strut with
$a = 0.20\,\si{\meter}$, $m = 1.2\,\si{\meter\squared\per\second}$, in
a tunnel speed $U_\infty = 1.5\,\si{\meter\per\second}$. The group
$m/(\pi U_\infty a) = 1.2/(\pi\times 1.5\times 0.20) = 1.27$, so the
half-length is $L = 0.20\sqrt{1 + 1.27} = 0.20\times 1.507 =
0.30\,\si{\meter}$, an oval $0.60\,\si{\meter}$ long. The half-thickness
$h$ is the value of $y$ where the dividing streamline $\psi = 0$ crosses
the $y$-axis, found from $0 = U_\infty h + (m/2\pi)(\theta_1 -
\theta_2)$ with $\theta_1 - \theta_2 = 2\arctan(a/h)$ at $x = 0$, a
transcendental relation
\[
h = \frac{m}{\pi U_\infty}\arctan\!\frac{a}{h},
\]
which iterates quickly. Starting from $h = 0.15\,\si{\meter}$, the right
side gives $(1.2/(\pi\times 1.5))\arctan(0.20/0.15) = 0.2546\times
0.927 = 0.236$; iterating to convergence lands near $h =
0.20\,\si{\meter}$, so the oval is about $0.60\,\si{\meter}$ long and
$0.40\,\si{\meter}$ thick, a blunt strut. The maximum surface speed at
the shoulder, from the superposed velocity field, runs about $1.3\,
U_\infty$ for this fineness ratio, giving a shoulder suction by
Bernoulli of $p_\text{min} - p_\infty = -\tfrac12\rho(1.3^2 - 1)
U_\infty^2 = -0.5\times 1000\times 0.69\times 1.5^2 =
-0.78\,\text{kPa}$. A slimmer strut, wanted for low disturbance, comes
from spreading the source and sink apart (larger $a$ at fixed $m$),
which lengthens the oval and drops the shoulder speed toward the
freestream; pulling them together at fixed product $m\cdot 2a$ recovers
the doublet and the circular cylinder of the previous section, with its
$2U_\infty$ shoulder speed. The oval interpolates between the two, and
the designer reads fineness ratio straight off the source-sink spacing.

\subsection{The thin-airfoil camber line as a vortex sheet}

Thin-airfoil theory replaces the airfoil by a sheet of vorticity laid
along the camber line, and the lift-slope result quoted earlier in
this section is the leading term of that construction. Carrying the
first two Fourier coefficients shows where camber enters and gives the
zero-lift angle, the single number that distinguishes a cambered
section from a flat plate.

\paragraph{Worked example: lift of a circular-arc cambered section.}
Model the airfoil by a bound vortex sheet of strength $\gamma(x)$
along the chord $0 \le x \le c$, with the flow-tangency condition that
the sheet's induced normal velocity cancels the freestream component
normal to the camber line at every chord station. Writing $x =
(c/2)(1 - \cos\theta)$ to map the chord onto $0 \le \theta \le \pi$
and expanding the sheet strength in the standard Fourier series, the
lift coefficient comes out as
\[
C_L = 2\pi\left(A_0 + \tfrac12 A_1\right),
\qquad
A_0 = \alpha - \frac{1}{\pi}\int_0^\pi \frac{dz_c}{dx}\,d\theta,
\quad
A_1 = \frac{2}{\pi}\int_0^\pi \frac{dz_c}{dx}\cos\theta\,d\theta,
\]
where $z_c(x)$ is the camber line and $dz_c/dx$ its slope. For a
symmetric section $dz_c/dx = 0$, the integrals vanish, $A_0 = \alpha$,
$A_1 = 0$, and $C_L = 2\pi\alpha$ recovers the flat-plate slope. For a
circular-arc camber of maximum ordinate $h$ at mid-chord, the slope is
$dz_c/dx = (4h/c)(1 - 2x/c) = (4h/c)\cos\theta$, so $A_0 = \alpha$ and
$A_1 = (2/\pi)\int_0^\pi (4h/c)\cos^2\theta\,d\theta = 4h/c$. The lift
coefficient is then
\[
C_L = 2\pi\alpha + 2\pi h/c,
\]
a flat-plate term plus a camber term that lifts the whole curve up by
a fixed amount. The \engterm{zero-lift angle} is the negative angle at
which the two cancel, $\alpha_{L=0} = -h/c$ (in radians). A section
with two percent camber ($h/c = 0.02$) carries $\alpha_{L=0} =
-0.02\,\text{rad} = -1.15^\circ$, meaning it still lifts at zero
geometric angle of attack, $C_L = 2\pi\times 0.02 = 0.126$, which is
why cambered sections are chosen for wings that must lift efficiently
at low angle. The lift slope $dC_L/d\alpha = 2\pi$ is unchanged by
camber; camber shifts the curve, it does not steepen it. The same two
coefficients fix the moment about the quarter-chord, $C_{m,c/4} =
-(\pi/4)(A_1 - A_2)$, the result that the quarter-chord is the
aerodynamic centre of a thin section, but the lift needs only $A_0$
and $A_1$.

\subsection{Unsteady Bernoulli: the oscillating U-tube column}

The unsteady term $\partial\phi/\partial t$ that steady Bernoulli
discards is the entire physics of an oscillating liquid column, and
the U-tube is the cleanest case in which to see it produce a
frequency. The example pairs with the water-hammer case study later in
the chapter: both are unsteady-flow problems, but the U-tube oscillates
slowly under gravity while water hammer rings at the acoustic speed.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-oscillating-column.tex}

\paragraph{Worked example: the natural frequency of a U-tube
manometer.} A liquid fills a U-tube of uniform bore, the total length
of the liquid column being $L$ (Figure~\ref{fig:vol03ch11:oscillating-column}).
Displace one surface up by $z$ and the other falls by $z$, so the
restoring pressure imbalance is the weight of a column of height $2z$,
giving a restoring force per unit area $\rho g\,(2z)$. The whole
column of mass per unit area $\rho L$ accelerates as a rigid slug
because the bore is uniform and the fluid is incompressible; Newton's
second law on the column reads
\[
\rho L\,\frac{d^2 z}{dt^2} = -2\rho g\,z,
\]
which is the unsteady Bernoulli statement in slug form: the
$\partial\phi/\partial t$ term integrated along the column is exactly
$L\,\ddot z$, and it balances the gravitational head difference
$2gz$. The motion is simple harmonic,
\[
\frac{d^2 z}{dt^2} + \frac{2g}{L}\,z = 0,
\qquad
\omega = \sqrt{\frac{2g}{L}},
\qquad
T = 2\pi\sqrt{\frac{L}{2g}}.
\]
For a column of total length $L = 1.6\,\si{\meter}$, the angular
frequency is $\omega = \sqrt{2\times 9.81/1.6} = 3.50\,\si{\radian\per
\second}$ and the period is $T = 2\pi/3.50 = 1.80\,\si{\second}$. The
frequency depends only on the column length, not on the bore or the
density, the same way a pendulum period depends only on its length.
The inviscid prediction is an undamped oscillation; a real column
decays as viscosity dissipates the kinetic energy, slowly in a wide
tube and fast in a narrow one, but the frequency is set by the
inviscid gravity-inertia balance and is accurate while the amplitude
lasts. Steady Bernoulli, with the $\partial\phi/\partial t$ term
dropped, predicts no oscillation at all: it would have the column sit
wherever it is released. The unsteady term is not a correction here,
it is the phenomenon.

\subsection{Free-streamline flow: the contraction coefficient of a jet}

A jet leaving a sharp-edged opening contracts to a cross-section
smaller than the opening, and the contraction coefficient that the
draining-tank and orifice examples took as a measured number can be
derived for one geometry from inviscid free-streamline theory. The
case shows that a coefficient usually quoted from tables has an exact
value in a clean limit, and it closes the loop on where the discharge
coefficient comes from.

\paragraph{Worked example: the Borda mouthpiece, $C_c = 1/2$.} A
\engterm{Borda mouthpiece} is a short tube projecting inward into a
tank, so that the approaching flow must reverse to enter it and the
jet springs clear of the tube walls
(Figure~\ref{fig:vol03ch11:orifice-vena-contracta} shows the analogous
vena contracta of a plain orifice). Apply the momentum theorem to a
control volume bounded by the tank walls and the jet. On the tank
walls the pressure is hydrostatic; because the mouthpiece projects
inward, the only wall area where the pressure is \emph{not} balanced
by an equal pressure on the opposite wall is the annulus directly
opposite the opening, of area equal to the opening area $A_o$, carrying
the hydrostatic pressure $\rho g h$ at the depth $h$ of the opening.
The net pressure force driving the flow is therefore $\rho g h\,A_o$.
The momentum leaving is $\dot m\,u = (\rho A_j u)\,u = \rho A_j u^2$,
where $A_j = C_c A_o$ is the contracted jet area and $u = \sqrt{2gh}$
is the Torricelli jet speed. Equating force to momentum flux,
\[
\rho g h\,A_o = \rho A_j u^2 = \rho\,(C_c A_o)\,(2 g h),
\]
and the head and the areas cancel to leave
\[
C_c = \tfrac12.
\]
The contraction coefficient of a Borda mouthpiece is exactly one half:
the jet area is half the opening area, a result that follows from the
momentum balance alone with no empirical input. A plain sharp-edged
orifice in a flat wall contracts less, because the approaching flow
does not have to reverse as completely; its contraction coefficient is
about $0.61$ to $0.64$, between the Borda value and unity, and that
range is what the draining-tank and orifice examples carried as a
measured $C_d \approx 0.62$. The discharge coefficient that those
examples treated as a table value is, in this clean inviscid limit, a
calculable geometric number; viscosity moves it by only a few percent.

\subsection{Nozzle reaction: the thrust a hose puts on the hand}

A hose nozzle accelerates the flow and so must push back on whatever
holds it, the same momentum balance that sizes a pipe bend applied to a
contraction open to the atmosphere. The example pairs Bernoulli, which
fixes the exit speed, with the momentum theorem, which turns the speed
change into the reaction the firefighter feels, and it shows the
pressure term and the momentum term entering on the same footing.

\paragraph{Worked example: reaction force on a firefighting nozzle.} A
nozzle contracts a hose flow from inlet diameter $D_1 =
65\,\si{\milli\meter}$ to exit diameter $D_2 = 25\,\si{\milli\meter}$
and discharges to atmosphere; the flow rate is $Q =
12\,\text{L/s}$. The exit area is $A_2 = \tfrac{\pi}{4}(0.025)^2 =
4.91\times 10^{-4}\,\si{\meter}^2$, so the jet leaves at
\[
u_2 = \frac{Q}{A_2} = \frac{0.012}{4.91\times 10^{-4}}
= 24.4\,\si{\meter\per\second}.
\]
The inlet area is $A_1 = \tfrac{\pi}{4}(0.065)^2 = 3.32\times
10^{-3}\,\si{\meter}^2$, giving an inlet speed $u_1 = Q/A_1 =
3.61\,\si{\meter\per\second}$. Bernoulli from the hose interior to the
atmospheric jet fixes the gauge pressure the hose must supply at the
nozzle inlet,
\[
p_1 = \tfrac12\rho(u_2^2 - u_1^2)
= 0.5\times 1000\times(595 - 13)
= 291\,\text{kPa},
\]
about three bar. Now the momentum theorem on a control volume wrapping
the nozzle, with the atmosphere at gauge zero on the jet, balances the
inlet pressure force and the holding force $R$ against the change in
momentum flux,
\[
p_1 A_1 + R = \dot m\,(u_2 - u_1),
\qquad
\dot m = \rho Q = 12\,\si{\kilogram\per\second}.
\]
The momentum change is $\dot m(u_2 - u_1) = 12\times(24.4 - 3.61) =
250\,\si{\newton}$, while the inlet pressure force is $p_1 A_1 =
291\,000\times 3.32\times 10^{-3} = 966\,\si{\newton}$. The holding
force is therefore $R = 250 - 966 = -716\,\si{\newton}$: the negative
sign says the hand must pull back on the nozzle with $716\,\si{\newton}$,
roughly the weight of a $73\,\si{\kilogram}$ person, to keep it from
flying forward. The pressure that the hose delivers dominates the bare
jet-momentum term by nearly a factor of four, the same lesson the pipe
bend taught: a reaction force computed from $\dot m\,\Delta u$ alone
badly understates the load, because the static pressure pushing on the
inlet face is the larger part of it. This is why a high-flow line needs
two crew on the nozzle.

\subsection{The sluice gate: discharge under a vertical lift gate}

A sluice gate meters an open channel by passing the flow under a raised
plate; the contracted stream below the gate is a free-surface analogue
of the orifice vena contracta, and the discharge follows from Bernoulli
on the free-surface streamline closed by continuity. The case is
distinct from the spillway weir, which passes flow over a crest; here
the flow passes under a gate and the depth, not a head over a crest,
sets the rate.

\paragraph{Worked example: discharge under a canal sluice gate.} Water
in a rectangular canal of width $b = 3\,\si{\meter}$ stands at upstream
depth $y_1 = 2.4\,\si{\meter}$ behind a vertical gate lifted to a clear
opening $a = 0.40\,\si{\meter}$. The jet emerging under the gate
contracts to a depth $y_2 = C_c a$ with a contraction coefficient
$C_c \approx 0.61$ for a sharp-edged gate, so $y_2 = 0.61\times 0.40 =
0.244\,\si{\meter}$. Bernoulli on the free-surface streamline, with
atmospheric pressure at both surfaces and the velocity heads measured
from the channel bed, reads
\[
y_1 + \frac{u_1^2}{2g} = y_2 + \frac{u_2^2}{2g},
\]
and continuity in the rectangular channel sets $u_1 y_1 = u_2 y_2$, so
$u_1 = u_2\,(y_2/y_1)$. Substituting and solving for the contracted
speed,
\[
u_2 = \sqrt{\frac{2g\,(y_1 - y_2)}{1 - (y_2/y_1)^2}}
= \sqrt{\frac{2\times 9.81\times(2.4 - 0.244)}
{1 - (0.244/2.4)^2}}
= \sqrt{\frac{42.3}{0.9897}}
= 6.54\,\si{\meter\per\second}.
\]
The discharge per unit width is $q = u_2 y_2 = 6.54\times 0.244 =
1.60\,\si{\meter}^2/\si{\second}$, so the total flow is $Q = b q =
3\times 1.60 = 4.79\,\si{\meter\cubed\per\second}$. The upstream speed
is $u_1 = u_2 y_2/y_1 = 6.54\times 0.244/2.4 = 0.665\,\si{\meter\per
\second}$, small enough that neglecting the upstream velocity head would
have changed the answer by under a percent, so the field shortcut
$q \approx C_c a\sqrt{2g y_1}$ gives $0.61\times 0.40\times\sqrt{2\times
9.81\times 2.4} = 0.244\times 6.86 = 1.67\,\si{\meter}^2/\si{\second}$,
within five percent. The gate is a calibrated meter: read the upstream
depth and the gate opening, and the discharge follows. The same
Bernoulli-plus-continuity pairing that meters a pipe with a Venturi
meters a channel with a gate.

\subsection{The siphon: how high the crown can climb before it cavitates}

A siphon lifts liquid over a barrier higher than the source surface and
delivers it to a lower outlet, running on the head difference between
the two free surfaces with no pump. Bernoulli gives the delivery speed
and, applied to the crown, the lowest pressure in the line, which sets a
hard ceiling on how high the crown may rise before the liquid flashes to
vapour. The case is the cavitation limit of the failure section met in a
benign device.

\paragraph{Worked example: the maximum crown height of a water siphon.}
A siphon draws water from an open reservoir, over a crown a height $h_c$
above the reservoir surface, and discharges to atmosphere a height
$h_d = 1.5\,\si{\meter}$ below the reservoir surface. Bernoulli from the
reservoir surface (at rest, atmospheric) to the open outlet (atmospheric)
gives the delivery speed from the net fall,
\[
u = \sqrt{2 g\,h_d} = \sqrt{2\times 9.81\times 1.5}
= 5.42\,\si{\meter\per\second},
\]
the same for any crown height, because the crown does no net work on a
loss-free flow. The crown pressure is the constraint. Bernoulli from the
reservoir surface to the crown, where the speed has already reached the
full line speed $u$ and the elevation is $h_c$ above the reservoir, gives
the crown gauge pressure
\[
p_c = p_\text{atm} - \rho g\,h_c - \tfrac12\rho u^2.
\]
Writing absolute pressures and demanding $p_c \ge p_v$ (the water must
not flash to vapour) sets the ceiling on the crown height,
\[
h_c^{\max} = \frac{p_\text{atm} - p_v}{\rho g} - \frac{u^2}{2g}
= \frac{(101.3 - 2.3)\times 10^3}{1000\times 9.81}
- \frac{5.42^2}{2\times 9.81}
= 10.09 - 1.50 = 8.6\,\si{\meter}.
\]
The first term, about $10\,\si{\meter}$, is the familiar
water-barometer height, the static limit on suction lift; the velocity
head subtracts a further $1.5\,\si{\meter}$ because the moving crown
flow is already below the static stagnation pressure. A crown taller
than about $8.6\,\si{\meter}$ pulls the crown pressure below the vapour
pressure, the line cavitates, the column breaks, and the siphon stops.
Real siphons lose a further margin to dissolved-gas release, which comes
out of solution well above the vapour pressure, so the practical ceiling
is nearer $7\,\si{\meter}$. The siphon is the cavitation limit turned
into a design rule: there is a height past which suction lift simply
cannot reach, set by the atmosphere on one side and the vapour pressure
on the other.

\subsection{Vignette: the spin speed of a lawn sprinkler}

The rotating lawn sprinkler is the angular-momentum theorem made into a
toy, and it is worth one short pass because the free-running speed
follows from setting the net torque to zero, a balance that needs the
moment of the absolute velocity, not the relative one, and that is
exactly where the intuition usually slips.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-sprinkler-angular.tex}

A two-arm sprinkler of arm radius $R = 0.15\,\si{\meter}$ passes a total
flow $Q = 0.30\,\text{L/s}$, split equally between two nozzles each of
exit area $A_n = 8\,\si{\milli\meter}^2 = 8\times 10^{-6}\,\si{\meter}^2$
(Figure~\ref{fig:vol03ch11:sprinkler-angular}). Each nozzle ejects its
share at a speed relative to the arm of
\[
v_\text{rel} = \frac{Q/2}{A_n}
= \frac{1.5\times 10^{-4}}{8\times 10^{-6}}
= 18.75\,\si{\meter\per\second}.
\]
The nozzles point tangentially, so each jet leaves with a tangential
velocity, in the rotating arm's frame, of magnitude $v_\text{rel}$
opposing the spin. In the fixed laboratory frame the water's absolute
tangential velocity is $v_\text{rel} - \Omega R$, the jet speed reduced
by the speed the nozzle tip itself is already moving. The
angular-momentum theorem about the pivot, for steady spin, sets the net
applied torque equal to the net flux of absolute angular momentum
carried out: the inlet water arrives on the axis with zero moment arm,
so the torque the bearing must supply is
\[
M_\text{bearing} = \dot m\,R\,(v_\text{rel} - \Omega R),
\]
with $\dot m = \rho Q = 0.30\,\si{\kilogram\per\second}$ the total mass
flow. A real bearing carries a small friction torque; the idealised
free-running sprinkler, with frictionless bearings, spins up until the
applied torque vanishes, which happens when the absolute tangential
velocity of the ejected water reaches zero,
\[
v_\text{rel} - \Omega R = 0
\quad\Longrightarrow\quad
\Omega_\text{max} = \frac{v_\text{rel}}{R}
= \frac{18.75}{0.15} = 125\,\si{\radian\per\second},
\]
about $1200$ revolutions per minute. At that speed the water leaves with
no absolute swirl, the sprinkler does no work on it, and there is
nothing left to drive the spin faster. A real sprinkler turns far
slower, a few revolutions per minute, because bearing friction and the
aerodynamic drag of the arms balance the driving torque well below
$\Omega_\text{max}$; equating the driving torque $\dot m R\,(v_\text{rel}
- \Omega R)$ to a friction torque is how the working speed is found. The
inviscid limit gives the ceiling, the same role it played for the hose
and the sink tap, and the lesson is the same: the absolute velocity, not
the relative one, carries the momentum across the control surface, and
forgetting the $\Omega R$ correction predicts a torque that never falls
to zero and a sprinkler that accelerates without bound.

\subsection{Case study: sizing a Venturi flow-metering station}

A flow-metering station is the inviscid relations of this chapter
turned into a piece of plant: continuity sets the throat speed,
Bernoulli sets the differential the transducer reads, a discharge
coefficient corrects for the small viscous and contraction effects,
the cavitation margin sets a floor on the throat size, and the
permanent pressure loss sets a running energy cost that the meter
choice trades against capital. We carry one station from a measurement
requirement to a sized, checked design.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-metering-egl.tex}

\paragraph{The requirement.} A water line of inside diameter $D_1 =
150\,\si{\milli\meter}$ must carry a flow that ranges from $Q_\text{min}
= 15\,\text{L/s}$ to $Q_\text{max} = 60\,\text{L/s}$, and the flow is
to be metered with a differential-pressure transducer that resolves
cleanly between $2\,\text{kPa}$ and $80\,\text{kPa}$. The line runs at
a static pressure of $p_1 = 250\,\text{kPa}$ absolute; the water is at
$20\,^{\circ}\mathrm{C}$, vapour pressure $p_v = 2.3\,\text{kPa}$
absolute. The task is to choose the throat diameter, confirm the
differential stays inside the transducer band across the full flow
range, check the cavitation margin at the worst case, and estimate the
permanent pressure loss the station imposes.

\paragraph{Step 1: the differential as a function of throat size.} The
line area is $A_1 = \tfrac{\pi}{4}D_1^2 = \tfrac{\pi}{4}(0.15)^2 =
1.767\times 10^{-2}\,\si{\meter}^2$. For a Venturi the ideal
differential follows from continuity and Bernoulli,
\[
\Delta p = \tfrac12\rho\,u_1^2\!\left[\left(\frac{A_1}{A_2}\right)^2
- 1\right]
= \tfrac12\rho\!\left(\frac{Q}{A_1}\right)^2\!
\left[\beta^{-4} - 1\right],
\]
writing $\beta = D_2/D_1$ for the diameter ratio, so $A_1/A_2 =
\beta^{-2}$. The differential scales as $Q^2$, so the ratio of the
extreme differentials is fixed at $(Q_\text{max}/Q_\text{min})^2 =
4^2 = 16$ whatever the throat, and the transducer band ratio
$80/2 = 40$ comfortably contains it. The throat must place that factor
of sixteen inside the band.

\paragraph{Step 2: choosing the diameter ratio.} Try $\beta = 0.5$
($D_2 = 75\,\si{\milli\meter}$, $\beta^{-4} = 16$). At $Q_\text{max} =
60\,\text{L/s} = 0.060\,\si{\meter\cubed\per\second}$ the approach
speed is $u_1 = Q/A_1 = 0.060/1.767\times 10^{-2} =
3.40\,\si{\meter\per\second}$, and
\[
\Delta p_\text{max} = \tfrac12\times 1000\times 3.40^2\times(16 - 1)
= 500\times 11.56\times 15 = 86.7\,\text{kPa},
\]
just over the $80\,\text{kPa}$ ceiling. Open the throat slightly to
$\beta = 0.55$ ($D_2 = 82.5\,\si{\milli\meter}$, $\beta^{-4} =
10.9$): then $\Delta p_\text{max} = 500\times 11.56\times(10.9 - 1) =
57.2\,\text{kPa}$, inside the ceiling, and $\Delta p_\text{min}$ at
$15\,\text{L/s}$ is smaller by the factor $16$, $57.2/16 =
3.6\,\text{kPa}$, above the $2\,\text{kPa}$ floor. The diameter ratio
$\beta = 0.55$ holds the differential inside the transducer band
across the full flow range, so we fix $D_2 = 82.5\,\si{\milli\meter}$,
$A_2 = \tfrac{\pi}{4}(0.0825)^2 = 5.35\times 10^{-3}\,\si{\meter}^2$.

\paragraph{Step 3: the discharge coefficient and the calibration.} The
ideal relation overstates the flow because the real throat velocity
profile is non-uniform and a thin boundary layer forms; a classical
Venturi carries a discharge coefficient $C_d \approx 0.97$ to $0.98$
nearly independent of flow over the operating range, so the working
calibration is
\[
Q = C_d\,A_2\sqrt{\frac{2\,\Delta p}{\rho\,(1 - \beta^4)}}.
\]
At the top of the range, $\sqrt{2\times 57\,200/(1000\times(1 -
0.0915))} = \sqrt{125\,900} = 355$, and $Q = 0.97\times 5.35\times
10^{-3}\times 355\times \ldots$ recovers $60\,\text{L/s}$ to within the
coefficient's tolerance. The near-constancy of $C_d$ for a Venturi is
what makes it the reference meter; an orifice plate at the same
$\beta$ would carry $C_d \approx 0.61$ and a stronger Reynolds-number
dependence, trading metering simplicity for the permanent loss
weighed in Step~5. The throat Reynolds number at $Q_\text{max}$ is
$\mathrm{Re}_2 = \rho u_2 D_2/\mu$ with $u_2 = Q/A_2 = 0.060/5.35\times
10^{-3} = 11.2\,\si{\meter\per\second}$ and $\mu = 1.0\times
10^{-3}\,\si{\pascal\second}$, giving $\mathrm{Re}_2 = 1000\times
11.2\times 0.0825/10^{-3} = 9.2\times 10^5$, well into the range where
$C_d$ is flat.

\paragraph{Step 4: the cavitation check.} The throat is the
lowest-pressure point. At $Q_\text{max}$ the throat speed is $u_2 =
11.2\,\si{\meter\per\second}$ and the static pressure there is
\[
p_2 = p_1 - \Delta p_\text{max} = 250 - 57 = 193\,\text{kPa absolute}.
\]
The cavitation number on the throat dynamic pressure is
\[
\sigma = \frac{p_2 - p_v}{\tfrac12\rho u_2^2}
= \frac{(193 - 2.3)\times 10^3}{0.5\times 1000\times 11.2^2}
= \frac{190.7\times 10^3}{62.7\times 10^3} = 3.0,
\]
far above the inception range (a smooth Venturi throat tolerates
$\sigma$ down toward $1$), so the station does not cavitate at full
flow. The smooth Venturi contour earns this margin; the same flow
through a sharp orifice, with its vena-contracta speed higher than the
bore speed, would sit much closer to inception, the trap of the
orifice-cavitation worked example earlier in this section.

\paragraph{Step 5: the permanent pressure loss.} The throat
differential is recovered almost entirely in the Venturi's gentle
diffuser; what is not recovered is the permanent loss the station
charges the pump for as long as it runs
(Figure~\ref{fig:vol03ch11:metering-egl}). For a classical Venturi the
permanent loss is a small fraction of the measured differential,
roughly $0.1$ to $0.2$ of $\Delta p$ for a well-made unit, so at full
flow the running loss is about $0.15\times 57 \approx 9\,\text{kPa}$.
The hydraulic power that loss dissipates is $P = Q\,\Delta p_\text{loss}
= 0.060\times 9000 = 540\,\si{\watt}$, a continuous draw on the pump. An
orifice plate at the same $\beta$ would recover far less of its throat
dip and impose a permanent loss of perhaps $0.6$ to $0.8$ of the
differential, four to five times the Venturi's, costing a few kilowatts
continuously. Over a year of running the energy difference dwarfs the
Venturi's higher purchase price, which is the standard reason a
metering station that runs continuously specifies a Venturi over the
cheaper plate.

\paragraph{Verdict.} A Venturi of throat diameter $82.5\,\si{
\milli\meter}$ ($\beta = 0.55$) in the $150\,\si{\milli\meter}$ line
meters the full $15$ to $60\,\text{L/s}$ range with a differential of
$3.6$ to $57\,\text{kPa}$, inside the transducer band; it carries a
flat discharge coefficient near $0.97$, holds a cavitation number of
$3.0$ at full flow with margin to spare, and imposes a permanent loss
near $9\,\text{kPa}$ (about $540\,\si{\watt}$) at the top of the range.
Continuity, Bernoulli, the cavitation number, and the permanent-loss
fraction, every one of them an inviscid relation of this chapter
corrected by a single empirical coefficient, carry the station from a
requirement to a sized and checked design. The one quantity the
inviscid analysis cannot supply on its own, the discharge coefficient,
is the disciplined boundary between this chapter and the viscous
treatment of the next.

\subsection{Case study: lift and induced drag of a finite wing}

The thin-airfoil result of this section is two-dimensional: it gives
the lift of a wing section of infinite span, on which every spanwise
station carries the same circulation and no vorticity is shed. A real
wing is finite, its circulation falls to zero at the tips, and the
shed vorticity exacts a drag that the two-dimensional theory cannot
see. Lifting-line theory (Prandtl, 1918) supplies the correction, and
the result is an inviscid drag, present even in a perfectly inviscid
fluid, that dominates the drag budget of an aircraft in slow flight.
We carry one wing from its planform and weight to a lift coefficient,
an induced angle, an induced drag, and a comparison of two aspect
ratios that decides a design.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-finite-wing-planform.tex}

\paragraph{The wing and the requirement.} A light aircraft of mass
$1100\,\si{\kilogram}$ (weight $W = 1100\times 9.81 =
10.8\,\text{kN}$) cruises at $U_\infty = 55\,\si{\meter\per\second}$
at low altitude, air density $\rho = 1.2\,\si{\kilogram\per\meter
\cubed}$. Its wing has span $b = 11\,\si{\meter}$ and planform area
$S = 16\,\si{\meter}^2$, so the chord averages $\bar c = S/b =
1.45\,\si{\meter}$ and the \engterm{aspect ratio} is
\[
AR = \frac{b^2}{S} = \frac{11^2}{16} = 7.6.
\]
The task is to find the cruise lift coefficient, the induced angle of
attack the trailing vortices impose, the induced drag and its power
cost, and to weigh that cost against a higher-aspect-ratio wing of the
same area.

\paragraph{Step 1: the cruise lift coefficient.} In steady level
flight the wing lift equals the weight, so
\[
C_L = \frac{W}{\tfrac12\rho U_\infty^2 S}
= \frac{10\,800}{0.5\times 1.2\times 55^2\times 16}
= \frac{10\,800}{29\,040}
= 0.372.
\]
The wing works at $C_L \approx 0.37$ in cruise, a typical light-aircraft
value.

\paragraph{Step 2: the induced angle and the effective angle of
attack.} For the minimum-induced-drag case (an elliptic spanwise
circulation distribution) the downwash is uniform across the span and
the \engterm{induced angle of attack} is
\[
\alpha_i = \frac{C_L}{\pi\,AR}
= \frac{0.372}{\pi\times 7.6}
= 0.0156\,\text{rad}
= 0.89^\circ.
\]
The trailing vortices tilt the local relative wind down by about nine
tenths of a degree across the wing
(Figure~\ref{fig:vol03ch11:finite-wing-planform}). The section
therefore sees an effective angle of attack smaller than the geometric
one by $\alpha_i$, so to hold the same $C_L$ the finite wing must be
set at a larger geometric angle than an infinite wing of the same
section, by exactly $\alpha_i$. The lift slope of the finite wing is
correspondingly shallower than the two-dimensional $2\pi$: $a =
a_0/(1 + a_0/(\pi\,AR))$ with $a_0 = 2\pi$ gives $a = 2\pi/(1 +
2/7.6) = 2\pi/1.263 = 4.98$ per radian, against $2\pi = 6.28$ for the
infinite wing.

\paragraph{Step 3: the induced drag.} The lift vector, perpendicular
to the local relative wind, is tilted back by $\alpha_i$ from the
vertical, so a component $C_L\,\alpha_i$ points downstream: the
\engterm{induced drag coefficient}
\[
C_{D,i} = \frac{C_L^2}{\pi\,e\,AR},
\]
with the \engterm{span efficiency factor} $e$ equal to $1$ for the
ideal elliptic loading and about $0.85$ to $0.95$ for a real
rectangular or tapered wing. Taking $e = 0.9$,
\[
C_{D,i} = \frac{0.372^2}{\pi\times 0.9\times 7.6}
= \frac{0.1384}{21.5}
= 0.00644.
\]
The induced drag force is $D_i = C_{D,i}\,\tfrac12\rho U_\infty^2 S =
0.00644\times 29\,040 = 187\,\si{\newton}$, and the power to overcome
it is
\[
P_i = D_i\,U_\infty = 187\times 55 = 10.3\,\text{kW}.
\]
About ten kilowatts of the cruise power goes into the trailing vortex
sheet. None of this drag is viscous; it is the kinetic energy left
behind in the rotating wake, and it would persist in a fluid of zero
viscosity. The profile drag of the section (the viscous part, the
business of the next chapter) adds a roughly constant floor
$C_{D,0}\approx 0.010$, so at cruise the induced drag is a comparable
fraction of the total, and at low speed, where $C_L$ is large, the
induced part dominates.

\paragraph{Step 4: the aspect-ratio trade.} Induced drag falls as the
aspect ratio rises, since $C_{D,i}\propto C_L^2/AR$ and the cruise
$C_L$ is fixed by the weight. Re-span the same $16\,\si{\meter}^2$ wing
to $b = 16\,\si{\meter}$ (a glider-like planform), giving $AR =
16^2/16 = 16$ and, with the same $e = 0.9$,
\[
C_{D,i} = \frac{0.1384}{\pi\times 0.9\times 16} = 0.00306,
\]
less than half the stubby wing's value
(Figure~\ref{fig:vol03ch11:drag-polar}). The induced power drops to
$P_i = 0.00306\times 29\,040\times 55 = 4.9\,\text{kW}$, a saving of
about $5\,\text{kW}$, more than half the induced power, for the same
lifting area. The slender wing pays for this in structure, a longer
span carries a larger root bending moment and weighs more, and in
roll response, so a transport aircraft compromises near $AR = 9$ to
$11$ while a sailplane, which lives or dies by its glide ratio, runs
$AR = 20$ to $30$ and accepts the structural penalty. The decision is
the inviscid induced-drag relation weighed against the structural cost,
and the drag polar is the picture the designer reads it from.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-drag-polar.tex}

\paragraph{Verdict.} The finite wing at cruise carries $C_L = 0.37$,
feels an induced angle of $0.89^\circ$, and burns about
$10\,\text{kW}$ on induced drag at $AR = 7.6$; doubling the aspect
ratio at constant area halves the induced drag and saves about
$5\,\text{kW}$, at the price of a heavier and less responsive wing.
The whole calculation is inviscid: the induced drag is the energy
deposited in the trailing vortices, a drag that d'Alembert's paradox
does \emph{not} cancel because the wing is finite and sheds vorticity.
The two-dimensional thin-airfoil theory earlier in this section gives
the section lift; lifting-line theory adds the spanwise finiteness and
the drag that comes with it; only the profile-drag floor waits for the
viscous treatment of the next chapter. A designer who sizes a wing on
two-dimensional section data alone, and forgets the induced term,
underestimates the cruise drag by a large fraction and the climb and
slow-flight drag by more.

\begin{mastery}
\textbf{Prandtl's lifting-line integral equation.} The finite-wing
results above are the closed-form fruit of a single integral equation.
Model the wing as a \engterm{lifting line}, a bound vortex along the
span whose circulation $\Gamma(y)$ varies from a maximum at the centre
to zero at the tips $y = \pm b/2$. Helmholtz's theorems forbid a vortex
line from ending in the fluid, so wherever $\Gamma$ changes along the
span a trailing vortex of strength $-d\Gamma/dy\,\,dy$ is shed into the
wake. That sheet of trailing vortices induces, by the Biot-Savart law,
a downwash at spanwise station $y_0$,
\[
w(y_0) = \frac{1}{4\pi}\int_{-b/2}^{b/2}
\frac{d\Gamma/dy}{y_0 - y}\,dy,
\]
a Cauchy principal-value integral over the shed sheet. The downwash
tilts the local relative wind down by $\alpha_i(y_0) = w(y_0)/U_\infty$,
and the section lift at $y_0$, set by the local effective angle of
attack through the two-dimensional lift slope and tied to the local
circulation by Kutta-Joukowski $\rho U_\infty\Gamma(y_0)$, closes the
loop. Assembling the three relations gives Prandtl's
\engterm{monoplane equation}, an integro-differential equation for the
unknown distribution $\Gamma(y)$. Substituting $y = (b/2)\cos\theta$
and the Fourier series $\Gamma(\theta) = 2bU_\infty\sum_n A_n\sin
n\theta$ turns it into an algebraic system for the coefficients $A_n$.
The lift integrates to $C_L = \pi\,AR\,A_1$, depending on the first
coefficient alone, while the induced drag is $C_{D,i} = \pi\,AR\,\sum_n
n A_n^2 = (C_L^2/\pi AR)(1 + \delta)$ with $\delta = \sum_{n\ge 2}
n(A_n/A_1)^2 \ge 0$. The drag is least when every higher coefficient
vanishes, $\delta = 0$, which leaves $\Gamma(\theta) = 2bU_\infty A_1
\sin\theta$, the \engterm{elliptic distribution}: the spanwise loading
that minimises induced drag for a given lift is an ellipse, and the
span efficiency factor is $e = 1/(1 + \delta)$. The elliptic wing of
the Spitfire was Prandtl's theorem made into aluminium. The whole of
the finite-wing case study, the induced angle $C_L/\pi AR$, the
induced drag $C_L^2/\pi e AR$, the shallower lift slope, falls out of
this one integral equation in its elliptic special case; the general
case is a few terms of the Fourier series and is the foundation of
every wing-loading calculation. A fuller development is in Anderson's
treatment of incompressible finite wings
\cite[chap.~5]{text:anderson-aerodynamics}.
\end{mastery}

\section{Further applications: circulation, entrainment, and pressure loads}\pathtag{standard}

The examples that follow carry the same three relations, continuity,
Bernoulli, and the momentum theorem, onto configurations the earlier
sections set up but did not work: a body that lifts because it spins
rather than because it is cambered, a pump with no moving part, a
junction that must be restrained in two directions at once, a blunt
nose whose stagnation pressure sets a structural load, and a roof that
is lost to suction rather than to downforce. Each earns its place by
adding a mechanism the earlier examples did not isolate, and each ends
in a sized number a designer would carry forward.

\subsection{The Magnus effect: lift from spin}

The lifting cylinder of the potential-flow section carried its
circulation as a mathematical term added to the stream function. A
real cylinder can carry that circulation physically, by spinning: the
no-slip condition drags the surrounding fluid around with the surface,
and the entrained swirl is a bound circulation of the same kind that
the point vortex supplied. The transverse force that results is the
\engterm{Magnus effect}, and the Kutta-Joukowski theorem gives it
without any new physics.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-magnus-cylinder.tex}

\paragraph{Worked example: the sideways force on a spinning cylinder.}
A cylinder of radius $R = 40\,\si{\milli\meter}$ and length $\ell =
0.6\,\si{\meter}$ spins at $\Omega = 400\,\si{\radian\per\second}$ in
an airstream of $U_\infty = 25\,\si{\meter\per\second}$, air density
$\rho = 1.2\,\si{\kilogram\per\meter\cubed}$
(Figure~\ref{fig:vol03ch11:magnus-cylinder}). If the surface fully
entrains the fluid, the surface speed is $\Omega R = 400\times 0.04 =
16\,\si{\meter\per\second}$, and the bound circulation is the surface
speed integrated once around the circumference,
\[
\Gamma = 2\pi R\,(\Omega R) = 2\pi\,\Omega R^2
= 2\pi\times 400\times 0.04^2
= 4.02\,\si{\meter\squared\per\second}.
\]
The Kutta-Joukowski lift per unit length is $L' = \rho U_\infty\Gamma =
1.2\times 25\times 4.02 = 121\,\si{\newton\per\meter}$, and over the
length the transverse force is $L = L'\ell = 121\times 0.6 =
72\,\si{\newton}$. The ratio of the surface speed to the freestream is
the \engterm{spin ratio} $\Omega R/U_\infty = 16/25 = 0.64$; the
idealised lift coefficient based on the frontal projection is
$C_L = L'/(\tfrac12\rho U_\infty^2\,2R) = 121/(0.5\times 1.2\times
625\times 0.08) = 4.0$, far above the stall-limited $C_L$ of a fixed
airfoil, which is one reason the Flettner rotor ship and the spinning
cylinder appear in high-lift schemes. Real rotating cylinders reach
only a fraction of the full-entrainment circulation because the
boundary layer does not turn the outer flow completely, so measured
$C_L$ saturates near $1$ to $2$ over the practical spin-ratio range;
the inviscid estimate is the ceiling the spin is chasing, and the
gap between it and the measured value is the boundary-layer story of
the next chapter. The same $\rho U_\infty\Gamma$ curves a sliced golf
ball and dips a topspin tennis shot, with the spin supplying the
circulation the camber supplies on a wing.

\subsection{The trailing-vortex pair behind a wing}

The starting-vortex argument of the vorticity section left the wing
carrying a bound circulation that must, by Helmholtz's theorems, trail
off the two tips as a pair of counter-rotating line vortices. Those
trailing vortices carry each other downward at a speed the point-vortex
field of the potential-flow section gives directly, and the descent
speed is the number a wake-separation rule turns into a following
distance. The example is worth carrying because it is the point vortex
put to a hazard calculation with no new machinery.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-vortex-pair.tex}

\paragraph{Worked example: how fast a wake descends behind a
transport.} A transport of weight $W = 2.3\times 10^6\,\si{\newton}$
and span $b = 60\,\si{\meter}$ cruises at $U_\infty =
250\,\si{\meter\per\second}$ in air of density $\rho =
0.41\,\si{\kilogram\per\meter\cubed}$
(Figure~\ref{fig:vol03ch11:vortex-pair}). For an elliptically loaded
wing the two trailing vortices behave as a pair separated by $d =
(\pi/4)b = 47\,\si{\meter}$, each of circulation equal to the wing's
root bound circulation. Lift equals weight, $W = \rho U_\infty\Gamma\,
b_\text{eff}$ with the effective span $b_\text{eff} = (\pi/4)b = d$, so
the shed circulation is
\[
\Gamma = \frac{W}{\rho U_\infty\,d}
= \frac{2.3\times 10^6}{0.41\times 250\times 47}
= 478\,\si{\meter\squared\per\second}.
\]
Each vortex sits in the velocity the other induces, $V = \Gamma/(2\pi
d)$, and the pair descends together at
\[
V = \frac{\Gamma}{2\pi d}
= \frac{478}{2\pi\times 47}
= 1.62\,\si{\meter\per\second}.
\]
The wake sinks at about a metre and a half per second, so in the two
minutes before it decays it descends roughly $190\,\si{\meter}$ below
the flight path, which is why arrival and departure separations put a
following light aircraft above, and well behind, the track of a heavy
one. The circulation scales with weight and inversely with speed and
span, so a heavy aircraft flying slow on approach trails the strongest
and slowest-sinking wake, the worst case the separation minima are
written for. The whole estimate is the point-vortex velocity $\Gamma/
(2\pi r)$ evaluated at the partner's location, the same field that gave
the lifting cylinder its swirl.

\subsection{Reading a Venturi with a differential manometer}

The Venturi worked examples so far took the pressure difference as a
number in kilopascals. In the field the difference is read off a
column of liquid: a U-tube manometer bridges the inlet and throat taps,
and the height the manometer fluid stands is the raw measurement. The
conversion from a column height to a flow rate is worth carrying once,
because the manometer's own fluid density enters and a reader who uses
the flowing-fluid density in place of the difference of densities gets
the wrong flow.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-venturi-manometer.tex}

\paragraph{Worked example: flow from a mercury manometer deflection.} A
Venturi meter of inlet diameter $D_1 = 100\,\si{\milli\meter}$ and
throat diameter $D_2 = 50\,\si{\milli\meter}$ meters water, and a
mercury U-tube manometer across the taps reads a deflection of $\Delta
h = 250\,\si{\milli\meter}$ (Figure~\ref{fig:vol03ch11:venturi-manometer}).
The manometer relation converts the column height to a pressure
difference using the difference between the mercury density $\rho_m =
13\,600\,\si{\kilogram\per\meter\cubed}$ and the water density $\rho =
1000\,\si{\kilogram\per\meter\cubed}$, because the water column above
the mercury on each leg carries part of the head:
\[
p_1 - p_2 = (\rho_m - \rho)\,g\,\Delta h
= (13\,600 - 1000)\times 9.81\times 0.250
= 30\,900\,\text{Pa} \approx 30.9\,\text{kPa}.
\]
The diameter ratio is $\beta = D_2/D_1 = 0.5$, so the area ratio is
$A_1/A_2 = \beta^{-2} = 4$ and $(A_1/A_2)^2 = 16$. The ideal inlet
speed follows from the combined continuity-Bernoulli relation,
\[
u_1 = \sqrt{\frac{2(p_1 - p_2)}{\rho\,[(A_1/A_2)^2 - 1]}}
= \sqrt{\frac{2\times 30\,900}{1000\times 15}}
= \sqrt{4.12}
= 2.03\,\si{\meter\per\second},
\]
and with $A_1 = \tfrac{\pi}{4}(0.10)^2 = 7.85\times
10^{-3}\,\si{\meter}^2$ the ideal flow is $A_1 u_1 = 15.9\,\text{L/s}$.
Applying the Venturi discharge coefficient $C_d = 0.97$ gives the
working flow $Q = 0.97\times 15.9 = 15.5\,\text{L/s}$. Had the reader
used the mercury density alone, $\rho_m g\,\Delta h = 33.4\,\text{kPa}$,
the flow would come out about four percent high; the density difference
is the correction the water backfill imposes, small when the manometer
fluid is much denser than the flowing fluid and not negligible when it
is not. The manometer turns the invisible pressure difference of the
Bernoulli calculation into a column height on a scale, and the reduction
from height to flow is two multiplications and a square root.

\subsection{The jet aspirator: a pump with no moving part}

An accelerated jet runs at a low static pressure, and that low
pressure can do useful work: it draws a second stream into the flow
and carries it along. The device that exploits this, the
\engterm{aspirator} or eductor, is the laboratory vacuum source, the
carburettor, the steam-ejector condenser pump, and the irrigation
fertiliser injector. Bernoulli sets the suction; the momentum theorem
sets what fraction of the driving flow's energy survives the mixing.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-aspirator.tex}

\paragraph{Worked example: the suction of a water aspirator.} A water
aspirator drives mains water through a throat where the speed reaches
$u_t = 14\,\si{\meter\per\second}$; upstream in the wider supply line
the speed is $u_1 = 2\,\si{\meter\per\second}$ and the static pressure
is the mains value $p_1 = 300\,\text{kPa}$ absolute
(Figure~\ref{fig:vol03ch11:aspirator}). Bernoulli from the supply line
to the throat gives the throat static pressure,
\[
p_t = p_1 - \tfrac12\rho(u_t^2 - u_1^2)
= 300\,000 - 0.5\times 1000\times(196 - 4)
= 300\,000 - 96\,000
= 204\,\text{kPa absolute}.
\]
That throat pressure sits above atmospheric, so this throat does not
yet suck. Open the throat further, to reach $u_t =
26\,\si{\meter\per\second}$: then $p_t = 300\,000 - 0.5\times 1000
\times(676 - 4) = 300\,000 - 336\,000 = -36\,\text{kPa}$ absolute,
which is unphysical, meaning the flow cavitates before it gets there.
The realisable limit is the throat speed that drops $p_t$ to the
vapour pressure $p_v = 2.3\,\text{kPa}$ absolute,
\[
u_t^{\max} = \sqrt{u_1^2 + \frac{2(p_1 - p_v)}{\rho}}
= \sqrt{4 + \frac{2(300\,000 - 2300)}{1000}}
= \sqrt{4 + 595}
= 24.5\,\si{\meter\per\second}.
\]
At that throat speed the aspirator pulls the side port down to the
vapour pressure of water, an absolute pressure of about
$2.3\,\text{kPa}$, which is the deepest vacuum a cold-water aspirator
can reach and is why the laboratory water aspirator bottoms out near
the vapour pressure of its supply water rather than at zero. A warmer
supply raises $p_v$ and gives a poorer vacuum, the everyday reason a
water aspirator works better on a cold day. The device converts a
few hundred kilopascals of supply head into a suction bounded not by
the pump but by the liquid's own vapour pressure, the same cavitation
limit the siphon and the orifice met, here turned into the working
principle rather than the failure mode.

\subsection{The dividing tee: a two-component restraint}

The pipe-bend example turned the flow through a right angle at
constant area and found a restraint load dominated by the static
pressure. A dividing tee splits one stream into two, so continuity and
the momentum theorem must be carried on three faces at once, and the
restraint carries a load in two directions that a single-inlet balance
would miss.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-tee-junction.tex}

\paragraph{Worked example: the restraint on a dividing tee.} Water
enters the run of a tee at $u_1 = 3\,\si{\meter\per\second}$ in a
$150\,\si{\milli\meter}$ line at gauge pressure $p_1 = 250\,\text{kPa}$
(Figure~\ref{fig:vol03ch11:tee-junction}). Two thirds of the flow
carries straight through into a $125\,\si{\milli\meter}$ outlet and one
third turns $90^\circ$ into a $100\,\si{\milli\meter}$ branch. The
inlet area is $A_1 = \tfrac{\pi}{4}(0.15)^2 = 1.767\times
10^{-2}\,\si{\meter}^2$, so the total flow is $Q = A_1 u_1 =
5.30\times 10^{-2}\,\si{\meter\cubed\per\second}$ and the mass flow is
$\dot m_1 = \rho Q = 53.0\,\si{\kilogram\per\second}$. The run outlet
carries $\dot m_2 = \tfrac23\dot m_1 = 35.3\,\si{\kilogram\per\second}$
through $A_2 = \tfrac{\pi}{4}(0.125)^2 = 1.227\times
10^{-2}\,\si{\meter}^2$ at $u_2 = (\tfrac23 Q)/A_2 = 0.0353/0.01227 =
2.88\,\si{\meter\per\second}$; the branch carries $\dot m_3 =
\tfrac13\dot m_1 = 17.7\,\si{\kilogram\per\second}$ through $A_3 =
\tfrac{\pi}{4}(0.10)^2 = 7.85\times 10^{-3}\,\si{\meter}^2$ at $u_3 =
(\tfrac13 Q)/A_3 = 0.01767/0.00785 = 2.25\,\si{\meter\per\second}$.
Take the outlet pressures equal to the inlet pressure for a low-loss
tee, $p_2 = p_3 = p_1 = 250\,\text{kPa}$. With $x$ along the run and
$y$ along the branch, the $x$-momentum balance carries the two run
faces plus the branch's zero streamwise momentum,
\[
p_1 A_1 - p_2 A_2 + R_x = \dot m_2 u_2 - \dot m_1 u_1
= 35.3\times 2.88 - 53.0\times 3 = 102 - 159 = -57\,\si{\newton},
\]
so $R_x = -57 - p_1 A_1 + p_2 A_2 = -57 - (250\,000)(0.01767) +
(250\,000)(0.01227) = -57 - 4418 + 3068 = -1407\,\si{\newton}$. The
$y$-balance carries the branch face and the branch momentum,
\[
-p_3 A_3 + R_y = \dot m_3 u_3 - 0
= 17.7\times 2.25 = 39.8\,\si{\newton},
\]
so $R_y = 39.8 + p_3 A_3 = 39.8 + (250\,000)(0.00785) =
39.8 + 1963 = 2003\,\si{\newton}$. The restraint carries $R_x =
-1407\,\si{\newton}$ and $R_y = 2003\,\si{\newton}$, a resultant of
$\sqrt{1407^2 + 2003^2} = 2448\,\si{\newton}$ directed into the branch
quadrant. The pressure terms dominate both components, the same lesson
the bend taught, and the transverse component is the one a naive
straight-through analysis would drop entirely: a tee walks sideways
off its supports under the branch reaction, so its thrust block must
resist the resultant, not just the streamwise part.

\subsection{Stagnation pressure as a structural load}

A Pitot tube reads the stagnation pressure to infer a speed; the same
stagnation pressure is a mechanical load wherever a flow is brought to
rest against a solid body, and a designer sizing a leading edge, a
bridge pier, or a vehicle's frontal panel needs it as a force, not as
an instrument reading. The stagnation streamline is the one streamline
that ends on the body at the front stagnation point, and Bernoulli
along it converts all of the freestream dynamic pressure into a
pressure rise on the surface.

\paragraph{Worked example: the stagnation load on a train nose in a
tunnel.} A train enters a tunnel at $U = 60\,\si{\meter\per\second}$
(about $216$ kilometres per hour). Relative to the train the tunnel
air approaches at $U$ and stagnates on the nose, so the stagnation
gauge pressure on the nose is the dynamic pressure,
\[
p_0 - p_\infty = \tfrac12\rho U^2 = 0.5\times 1.2\times 60^2
= 2160\,\text{Pa} \approx 2.2\,\text{kPa}.
\]
Over a nose frontal area of about $10\,\si{\meter}^2$ the leading
stagnation region carries a net forward load of order $p_0 A =
2160\times 10 = 21.6\,\si{\kilo\newton}$, though the load falls away
from the stagnation point as $C_p$ drops from $1$, so the integrated
drag contribution is a fraction of this. The stagnation number matters
more as a pressure the glazing and body panels must hold: a windscreen
of area $2\,\si{\meter}^2$ near the stagnation region carries about
$2160\times 2 = 4.3\,\si{\kilo\newton}$, a load the mounting must
carry without deflection that would break the seal. The same
$\tfrac12\rho U^2$ scaling is why a train passing a platform at speed
produces a pressure pulse that can knock a person off balance: the
stagnation and suction fields sweep past at the train speed, and at
$60\,\si{\meter\per\second}$ the peak-to-peak swing is a few
kilopascals over a fraction of a second. Doubling the speed quadruples
every one of these loads, the quadratic scaling that makes
high-speed-rail aerodynamic loads grow faster than the speed itself.

\subsection{Wind uplift on a roof: suction, not downforce}

Wind over a building accelerates across the roof, and by Bernoulli the
external pressure there falls into suction. A roof is lost in a storm
by lifting off, and sizing its fixings is an inviscid pressure
calculation on the external flow paired with a statement about the
internal pressure the building holds.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-roof-pressure.tex}

\paragraph{Worked example: the uplift on a flat roof in a gale.} A
storm wind of $U = 30\,\si{\meter\per\second}$ (about $108$ kilometres
per hour, a strong gale) blows over a flat-roofed building of plan
area $A_\text{roof} = 200\,\si{\meter}^2$, air density $\rho =
1.2\,\si{\kilogram\per\meter\cubed}$
(Figure~\ref{fig:vol03ch11:roof-pressure}). The dynamic pressure is
$q = \tfrac12\rho U^2 = 0.5\times 1.2\times 900 = 540\,\text{Pa}$. Over
a flat roof the external pressure coefficient in the accelerated flow
is a suction, typically $C_{p,e} \approx -0.6$ to $-1.0$ averaged over
the roof, with worse peaks at the windward edge; take $C_{p,e} =
-0.7$, so the external gauge pressure on the roof is $p_e = C_{p,e}\,q
= -0.7\times 540 = -378\,\text{Pa}$. If a windward door or a broken
window leaves the interior at the windward stagnation pressure, the
internal coefficient is $C_{p,i} \approx +0.2$, so $p_i = +0.2\times
540 = +108\,\text{Pa}$. The net uplift per unit area is the difference,
\[
\Delta p = p_i - p_e = 108 - (-378) = 486\,\text{Pa},
\]
and the total uplift is $\Delta p\,A_\text{roof} = 486\times 200 =
97\,\si{\kilo\newton}$. A lightweight roof deck weighing perhaps
$0.4\,\si{\kilo\newton}$ per square metre carries a dead load of only
$0.4\times 200 = 80\,\si{\kilo\newton}$: the uplift exceeds the dead
weight, and the roof lifts unless it is anchored to the structure
below. This is why building codes size roof fixings for suction and
why a building with a windward opening is at greater risk than a
sealed one, the internal pressure adding to the external suction
rather than balancing it. The wind never pushes the roof down; it
sucks it up, and the sign of that load is the whole of the design
lesson.

\begin{estimation}
\textbf{Estimate the speed a headwind must reach before it stops a
cyclist who is putting out a steady effort.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A cyclist on the flat at a steady moderate effort holds perhaps $P
\approx 150\,\si{\watt}$ against aerodynamic drag, which dominates
above walking pace. The drag force is $D = C_D A\,\tfrac12\rho v_r^2$
with $v_r$ the air speed relative to the rider, a drag-area $C_D A
\approx 0.4\,\si{\meter}^2$ for a rider on the hoods, and air density
$\rho = 1.2\,\si{\kilogram\per\meter\cubed}$. On still air the rider's
sustainable ground speed $v_0$ satisfies $P = D v_0 = C_D A\,\tfrac12
\rho v_0^3$, giving $v_0^3 = 150/(0.4\times 0.5\times 1.2) = 625$, so
$v_0 = 8.6\,\si{\meter\per\second}$, about $31$ kilometres per hour, a
sensible figure.

Into a headwind of speed $w$, the air speed is $v_r = v_g + w$ with
$v_g$ the ground speed, and the same power balance is $P = C_D A\,
\tfrac12\rho(v_g + w)^2\,v_g$. The rider is stopped ($v_g \to 0$) when
even at zero ground speed the wind alone would demand more than the
available power to make any headway, but the sharper everyday marker is
where the ground speed collapses. Setting $v_g$ small, the power goes
almost entirely into fighting the wind's dynamic pressure over the
frontal area: the rider crawls once $\tfrac12\rho w^2\,C_D A$ times a
walking $v_g$ absorbs the $150\,\si{\watt}$. Taking $v_g \approx
1.5\,\si{\meter\per\second}$ (a walk) and solving $150 = 0.4\times 0.5
\times 1.2\times(1.5 + w)^2\times 1.5$ gives $(1.5 + w)^2 = 417$, so
$w \approx 19\,\si{\meter\per\second}$, about $68$ kilometres per hour.

\smallskip
The postmortem: a headwind near $20\,\si{\meter\per\second}$ reduces a
strong-gale-day rider to a walking crawl, and the quadratic dynamic
pressure is why. The estimate rests on Bernoulli's $\tfrac12\rho v^2$
turned into a force through a drag area, the same dynamic pressure that
lifts the roof and loads the train nose. The largest uncertainty is
the drag area, which a rider changes by a factor of two between an
upright and a tucked posture, which is exactly why a rider tucks into
a headwind.
\end{estimation}

\subsection{Vignette: the fertigation injector}

A field-irrigation system that must add fertiliser to the water stream
faces a choice between a powered dosing pump and a device with no
moving part, and the inviscid aspirator supplies the second. The
\engterm{Venturi injector} used in agriculture (a proprietary line of
them is sold under the Mazzei name) is a moulded Venturi with a suction
port at the throat: the irrigation flow through the throat drops the
local pressure below the pressure in a fertiliser tank vented to
atmosphere, and the pressure difference drives the concentrate up a
feed line into the throat, where it mixes and carries downstream to the
field. The device is the aspirator of the last section turned into a
line component.

The design case is a differential requirement. The throat must run at
a pressure low enough that the fertiliser tank, sitting at atmospheric
pressure plus a small static head, can push concentrate into it against
the feed-line loss. For an irrigation main at $250\,\text{kPa}$ gauge
carrying water at a line speed of $2\,\si{\meter\per\second}$, a throat
sized to reach $14\,\si{\meter\per\second}$ drops the throat pressure by
$\tfrac12\rho(u_t^2 - u_1^2) = 0.5\times 1000\times(196 - 4) =
96\,\text{kPa}$, to a throat gauge pressure of $250 - 96 =
154\,\text{kPa}$, still well above atmospheric. This throat does not
suck, because the line pressure is high. The working injector runs the
throat much faster, or accepts a large pressure drop across the whole
unit, so that the throat pressure falls below atmospheric and the tank
can feed it; a typical unit is specified by a minimum inlet-to-outlet
pressure differential, often a factor of about two across the device,
below which the suction stops. The trade is stark and is the reason the
injector is chosen or rejected: the device has no moving part, no
electrical supply, and no maintenance, but it charges a permanent
pressure loss of a third to a half of the inlet pressure, paid by the
pump for as long as it runs. Where the irrigation pump has head to
spare, the injector is the simplest possible metering of a second
fluid into a stream; where head is scarce, a powered dosing pump that
adds concentrate without a throat loss wins. The whole selection is the
throat Bernoulli balance of the aspirator weighed against the permanent
loss of the Venturi station, the two inviscid results of this chapter
meeting in one field component.

\subsection{Vignette: the total-energy variometer}

A glider pilot needs to know whether the air the sailplane is flying
through is rising or sinking, and a plain altimeter cannot tell that
apart from the climb and dive the pilot commands with the stick. The
instrument that separates the two, the \engterm{total-energy
variometer}, is a direct reading of the unsteady Bernoulli sum, and it
is worth a short pass because it turns the abstract statement ``the
Bernoulli constant is the mechanical energy of the air'' into a needle
a pilot watches.

A plain variometer measures the rate of change of static pressure,
that is, the rate of climb or descent. The trouble is that a pilot who
pulls up trades airspeed for height: the glider climbs, the plain
variometer shows a climb, but the air itself may be sinking and the
glider is merely coasting up its own kinetic energy, about to pay it
back. The total-energy variometer corrects for this by measuring the
rate of change of the \emph{total} head, the sum of the pressure head
and the velocity head, rather than the pressure head alone. A pull-up
that converts speed to height leaves the total head unchanged, so the
total-energy variometer reads zero, while a genuine updraft that adds
energy to the glider from the rising air moves the needle. The
correction is applied with a small venturi or a shaped
\engterm{total-energy probe} mounted in the airstream: the probe is
designed so that its port pressure is the static pressure minus the
dynamic pressure, $p - \tfrac12\rho u^2$, and connecting the
variometer's reference to that port makes the instrument respond to
$d(p + \tfrac12\rho u^2)/dt$, the rate of change of the Bernoulli sum,
rather than to $dp/dt$ alone. The probe is a Bernoulli device sized so
that its recovery exactly cancels the airspeed term the pilot can
command, leaving only the energy the atmosphere adds or removes. A
glider pilot reading a total-energy variometer is reading the time
derivative of the Bernoulli constant of the surrounding air, and the
whole of soaring flight, staying aloft by finding air that carries more
mechanical energy than the glider is spending, is the pilot chasing a
positive reading on that instrument. The steady Bernoulli equation of
this chapter says the constant is fixed along a streamline; the
variometer is the tool that reports when the glider crosses into a
streamline carrying a larger one.

\subsection{Vignette: the centrifugal pump and its cavitation floor}

The angular-momentum theorem gave the Euler turbomachine equation as a
raw relation between shaft torque and the change in swirl; a
centrifugal pump is that equation turned into a machine, and its two
inviscid results, the head it can raise and the suction limit at its
inlet, are the whole of a first-pass selection. The pump is a fresh
device for this chapter, and it brings the swirl side of the
angular-momentum theorem and the cavitation limit of the failure
section together in one component.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-pump-head-flow.tex}

A centrifugal impeller of outer radius $r_2$ spinning at angular speed
$\Omega$ flings the fluid outward, adding swirl. The Euler turbomachine
equation, with no swirl at the eye ($u_{t,1} = 0$) and a tip blade
speed $U_2 = \Omega r_2$, gives the ideal specific work as $w =
U_2\,u_{t,2}$, and dividing by $g$ gives the \engterm{Euler head}
$H = U_2\,u_{t,2}/g$. For a blade that throws the fluid straight out
radially the exit swirl equals the tip speed, $u_{t,2} = U_2$, and the
shut-off head (zero flow) is $H_0 = U_2^2/g$. Take an impeller of
outer diameter $0.30\,\si{\meter}$ ($r_2 = 0.15\,\si{\meter}$) turning
at $\Omega = 300\,\si{\radian\per\second}$ (about $2900$ revolutions
per minute): the tip speed is $U_2 = \Omega r_2 = 300\times 0.15 =
45\,\si{\meter\per\second}$, and the shut-off head is $H_0 = 45^2/9.81
= 206\,\si{\meter}$. Backward-curved blades, the usual choice, reduce
the exit swirl below the tip speed and drop the head to perhaps $60$ to
$80\,\%$ of that, so the working head is near $130$ to
$165\,\si{\meter}$, and it falls with flow along the drooping curve of
Figure~\ref{fig:vol03ch11:pump-head-flow}. The head the machine can
raise is the tip speed squared over $g$, corrected downward by the
blade angle and the internal loss; the ideal Euler line is the ceiling,
and slip and friction are the gap to the real curve, the same
energy-not-conserved bookkeeping the sudden expansion carried.

The inlet is where Bernoulli sets a hard floor. As the fluid
accelerates into the impeller eye its static pressure falls, and if it
falls to the vapour pressure the pump cavitates, the impeller pits, and
the head collapses. The margin is the \engterm{net positive suction
head} available,
\[
\text{NPSH}_a = \frac{p_\text{inlet} - p_v}{\rho g} + \frac{u_1^2}{2g},
\]
the head by which the inlet stagnation pressure exceeds the vapour
pressure. The pump carries a required value $\text{NPSH}_r$ set by the
worst suction on the blade, and the selection rule is $\text{NPSH}_a >
\text{NPSH}_r$ with margin. Suppose the pump lifts water from an open
well $4\,\si{\meter}$ below its inlet, so the inlet gauge pressure is
already $-\rho g\times 4 = -39\,\text{kPa}$, an absolute pressure of
$101 - 39 = 62\,\text{kPa}$ before any velocity head. With an inlet
speed $u_1 = 2\,\si{\meter\per\second}$ the available suction head is
\[
\text{NPSH}_a = \frac{(62 - 2.3)\times 10^3}{1000\times 9.81}
+ \frac{2^2}{2\times 9.81}
= 6.09 + 0.20 = 6.3\,\si{\meter}.
\]
A pump requiring $\text{NPSH}_r = 4\,\si{\meter}$ runs with a
$2.3\,\si{\meter}$ margin and does not cavitate; the same pump asked to
lift from $7\,\si{\meter}$ below its inlet loses another
$3\,\si{\meter}$ of head and drops below its requirement, so it
cavitates and must be lowered nearer the water or replaced with a
lower-$\text{NPSH}_r$ design. The suction lift is bounded by the same
atmosphere-minus-vapour-pressure ceiling the siphon and the aspirator
met, here the number that decides where the pump may sit. The two
inviscid results, the Euler head at the outlet and the suction floor at
the inlet, size and place the machine before a single loss coefficient
is looked up.

\begin{mastery}
\textbf{Added mass: the fluid a body must accelerate with it.} An
inviscid fluid exerts no steady drag on a body (d'Alembert), but it
does resist a body that \emph{accelerates}, because setting the body in
motion sets the surrounding fluid in motion too, and that fluid carries
kinetic energy the accelerating force must supply. The effect is
\engterm{added mass} (or virtual mass), and it is a purely inviscid,
purely potential-flow phenomenon. For a sphere of radius $R$
accelerating through a fluid of density $\rho$, the velocity potential
of the disturbance flow is a doublet, and integrating the fluid kinetic
energy $\tfrac12\rho\int|\nabla\phi|^2\,dV$ over the exterior gives an
energy $\tfrac12(m_a)U^2$ with the added mass
\[
m_a = \tfrac12\rho\,\tfrac43\pi R^3,
\]
exactly half the mass of the displaced fluid. The equation of motion
of the sphere accelerating under an applied force $F$ in the fluid is
then $(m + m_a)\dot U = F$: the body behaves as if it carried an extra
mass $m_a$. For a long cylinder the added mass per unit length is the
full displaced fluid, $\rho\pi R^2$; for a flat plate accelerating
broadside it is $\rho\pi(b/2)^2$ per unit length with $b$ the plate
width. Added mass is why a balloon released underwater accelerates
upward far more slowly than its buoyancy-to-weight ratio alone would
predict, why a ship's hull feels heavier to accelerate than its own
mass, and why the natural frequency of a structure vibrating in water
is lower than in air. It also carries the transient of the U-tube
column and the surge tank: the ``mass'' that oscillates includes the
fluid the moving surface must drag with it. The number falls out of the
same potential-flow machinery as the lifting cylinder, with the
disturbance potential integrated for its energy rather than
differentiated for its velocity, and it is the one force an inviscid
fluid exerts that survives into every real unsteady flow, undiminished
by viscosity because it is an inertial effect, not a frictional one.
\end{mastery}

\begin{mastery}
\textbf{The Rankine combined vortex and the tornado core.} The point
vortex of the potential-flow section has a tangential velocity
$u_\theta = \Gamma/(2\pi r)$ that grows without bound as $r\to 0$, an
unphysical singularity at the centre. A real concentrated vortex, a
bath drain, a tornado, a wingtip vortex, resolves the singularity with
a rotational core, and the simplest model that captures both regions is
the \engterm{Rankine combined vortex}: a forced-vortex core in
solid-body rotation surrounded by a free-vortex potential flow. Inside
the core radius $r_c$ the fluid rotates as a rigid body, $u_\theta =
\Omega r$, with uniform vorticity $\omega = 2\Omega$; outside, the flow
is the irrotational free vortex $u_\theta = \Omega r_c^2/r$, with zero
vorticity. The two match in speed at $r = r_c$, where the tangential
velocity peaks at $u_{\theta,\max} = \Omega r_c$. The pressure field
follows from the two balances this chapter has already used
separately. In the irrotational region Bernoulli holds across
streamlines (the flow is irrotational, so the constant is field-wide),
and integrating $dp/dr = \rho u_\theta^2/r$ inward gives a pressure
that falls as $r$ shrinks, $p(r) = p_\infty - \tfrac12\rho\,
u_{\theta,\max}^2\,(r_c/r)^2$. In the rotational core Bernoulli does
\emph{not} hold across streamlines (the core is rotational, the
Crocco/Bernoulli-constant-varies warning of the vorticity section), so
the pressure is found instead from the forced-vortex balance $dp/dr =
\rho\Omega^2 r$, giving a pressure that continues to fall to the axis,
reaching $p_\text{axis} = p_\infty - \rho\,u_{\theta,\max}^2$, twice as
deep below ambient as the core edge. The factor of two is the signature
of the combined vortex: half the pressure drop is in the outer
irrotational sweep, half in the rotational core, and a barometer
carried into a tornado records both. The model is the working
first-order description of every concentrated vortex, and it is exactly
the boundary the failure section draws, made quantitative: Bernoulli
carries across streamlines in the irrotational annulus and fails across
them in the rotational core, and the pressure field must be assembled
from the two balances, matched at the core radius, precisely because
the flow changes its rotational character there.
\end{mastery}

\section{Failure: where Bernoulli does not apply, and why students misuse it}\pathtag{core}

Every chapter in this work has a failure section. The failure mode
for inviscid analysis is qualitatively different from the failure
modes of structural mechanics: an inviscid Bernoulli calculation
applied to a configuration that violates its assumptions gives a
plausible-looking number that is wrong, sometimes by an order of
magnitude.

\subsection{Bernoulli across a hydraulic jump}

Water flowing rapidly down a spillway and entering a slower
tailrace transitions through a \engterm{hydraulic jump}: an abrupt
rise in water depth from a supercritical fast-shallow flow to a
subcritical slow-deep flow. Bernoulli applied across the jump
predicts a pressure rise consistent with the velocity decrease,
but the actual energy loss is substantial: the jump dissipates a
fraction of the upstream kinetic energy as turbulence, heat, and
acoustic radiation. The fractional energy loss can exceed $50\,\%$
for strong jumps (Froude numbers above $5$).

The reason Bernoulli fails across the jump is that the flow is not
inviscid in the jump region: intense turbulent dissipation
converts mechanical energy to heat. The control-volume momentum
balance still holds (it makes no inviscid assumption), and the
correct analytical tool is the momentum equation across the jump.
The momentum analysis gives the downstream depth and the energy
loss; Bernoulli applied through the jump gives the wrong depth
and conserves an energy that is in fact being dissipated.

\subsection{Bernoulli inside a boundary layer}

A real flow past a solid surface develops a thin boundary layer
(chapter 12 of this volume) in which viscous stresses dominate.
Inside the boundary layer, the streamwise velocity drops from the
freestream value to zero at the wall. Bernoulli predicts that the
pressure must rise correspondingly along the streamline that
approaches the wall, by an amount $\tfrac{1}{2}\rho U_\infty^2$.

The actual pressure distribution at the wall is set by the
inviscid flow just outside the boundary layer, not by the
within-boundary-layer streamline. The boundary-layer approximation
(Prandtl, 1904; chapter 12 of this volume) states that the
pressure is constant across the boundary layer (in the wall-normal
direction): the boundary layer is so thin that the pressure
imposed by the outer inviscid flow penetrates to the wall
unchanged. A Bernoulli calculation along a streamline that crosses
into the boundary layer is therefore qualitatively wrong: the
pressure does not rise to the stagnation value, and the velocity
loss is due to viscous dissipation, not pressure recovery.

\subsection{The lift formula misused}

A common misconception holds that an airfoil generates lift
because air takes longer to travel over the longer-curved upper
surface than over the shorter lower surface, and ``must''
therefore travel faster on the upper surface to ``meet up'' with
the lower-surface flow at the trailing edge. The argument is the
\engterm{equal-transit-time fallacy}, and it is wrong on every
point: there is no physical principle that forces fluid particles
to meet up at the trailing edge (they do not), the upper-surface
flow on a real airfoil is much faster than the equal-transit
calculation predicts (often by a factor of two or more), and flat
plates and inverted cambered airfoils generate lift in
configurations where the equal-transit argument predicts no lift
or negative lift.

The correct explanation rests on the Kutta-Joukowski theorem and
the establishment of bound circulation through the Kutta
condition: a real airfoil at angle of attack develops a sharp
trailing-edge flow pattern that fixes the circulation at the value
required to keep the rear stagnation point at the trailing edge
itself (the \engterm{Kutta condition}, Martin Kutta, 1902). The
bound circulation, combined with the freestream, produces a
faster flow on the upper surface and a slower flow on the lower
surface; the pressure differential follows by Bernoulli; the
integrated pressure differential is the lift, equal to
$\rho U_\infty \Gamma$. The argument is closed once the
circulation is fixed by the Kutta condition.

\subsection{Bernoulli in unsteady or rotational flow}

A pulsating flow, a starting transient, or a flow with significant
vorticity along the streamline of interest all violate Bernoulli's
assumptions. The correct generalisation for unsteady irrotational
flow is the \engterm{unsteady Bernoulli equation}
\[
\frac{\partial \phi}{\partial t} + \tfrac{1}{2}u^2 + \frac{p}{\rho}
+ gz = f(t),
\]
which carries an additional time-derivative term. For a rotational
flow, the Bernoulli constant differs between streamlines, and the
relation cannot be applied across streamlines without an analysis
of the vorticity distribution. Students who memorise the
incompressible steady form and apply it to oscillating pipe flows,
starting jets, or vortex-shedding wakes get numerically incorrect
answers; the discipline is to check the five assumptions before
writing the equation.

\subsection{Cavitation: when Bernoulli predicts a pressure the
liquid cannot sustain}

A liquid cannot carry an absolute pressure below its vapour pressure
$p_v$; at that pressure it boils. Bernoulli is the calculation that
warns when a design approaches this limit. Where a flow accelerates
through a constriction, a pump inlet, a valve, a propeller-blade
suction surface, or the toe of a spillway, the static pressure drops
by $\tfrac12\rho(u_2^2 - u_1^2)$, and a high enough local speed drives
the absolute pressure below $p_v$. Vapour bubbles form, are swept
downstream into a higher-pressure region, and collapse. The collapse
is violent: the bubble walls accelerate inward and the implosion sends
a pressure spike of hundreds of megapascals into any nearby surface,
pitting bronze impellers, steel valve seats, and concrete spillway
faces over time. The phenomenon is \engterm{cavitation}, and its onset
is governed not by viscosity but by the inviscid Bernoulli pressure
drop.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-cavitation.tex}

Figure~\ref{fig:vol03ch11:cavitation} traces the static pressure
along a streamline through a constriction down past the vapour
pressure of water (about $2.3\,\text{kPa}$ absolute near
$20\,^{\circ}\mathrm{C}$). The dimensionless group that ranks
cavitation risk is the \engterm{cavitation number}
\[
\sigma = \frac{p - p_v}{\tfrac12\rho U^2},
\]
the margin between local static pressure and vapour pressure measured
in dynamic pressures; cavitation begins when $\sigma$ falls to a
geometry-dependent inception value. The lesson for the failure
section is that Bernoulli here gives the right answer for the wrong
comfort: the inviscid pressure prediction is accurate up to the point
of inception, and it is exactly that accurate prediction of a very low
pressure that flags the hazard. A designer who computes the throat
pressure and finds it below $p_v$ has been warned by Bernoulli; one
who never computes it learns about cavitation from the pitted hardware.

\subsection{Water hammer: when the steady assumption breaks
catastrophically}

Closing a valve at the end of a pipe carrying flowing liquid stops the
flow, and the moving column of liquid cannot stop instantly. The
kinetic energy of the column converts into a pressure surge that
travels back up the pipe as an acoustic wave at the speed of sound in
the confined liquid, reflects, and returns, ringing the pipe like a
struck bell. The phenomenon is \engterm{water hammer}, the audible
bang in household plumbing when a tap or appliance valve shuts
quickly. Steady Bernoulli has nothing to say about it: the flow is
violently unsteady, and the relevant term is the $\partial\phi/
\partial t$ contribution that steady Bernoulli discards.

The peak pressure rise for an instantaneous stop is the
\engterm{Joukowsky relation}
\[
\Delta p = \rho a\,\Delta u,
\]
where $a$ is the pressure-wave speed (close to the liquid's sonic
speed, reduced by pipe-wall elasticity) and $\Delta u$ is the change
in flow speed. For water with $a \approx 1200\,\si{\meter\per\second}$
and a flow brought to rest from only $2\,\si{\meter\per\second}$, the
surge is $\Delta p = 1000\times 1200\times 2 = 2.4\,\text{MPa}$, some
twenty-four atmospheres, far above the steady operating pressure and
enough to rupture fittings. The surge is bounded only if the valve
closes slowly relative to the wave's round-trip time $2L/a$ (with $L$
the pipe length); a slow closure lets successive reflections partly
cancel the surge. Surge tanks, air chambers, and pressure-relief
valves are the standard mitigations. Water hammer is the
unsteady-flow failure mode of this chapter made concrete: the steady
assumption is not approximately wrong, it discards the entire
physics.

\subsection{Case study: water-hammer surge in a hydroelectric
penstock}

The water-hammer failure mode deserves a full end-to-end treatment,
because the engineering decision it forces, how fast a valve may be
allowed to close, follows from the inviscid relations of this chapter
strung together with the elasticity of the pipe and the geometry of
the installation. We carry one penstock through from its steady
operating state to a verdict on its pressure rating.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-penstock-layout.tex}

\paragraph{The installation.} A small hydroelectric scheme
(Figure~\ref{fig:vol03ch11:penstock-layout}) draws from a reservoir
through a steel penstock of length $L = 600\,\si{\meter}$, inside
diameter $D = 1.2\,\si{\meter}$, and wall thickness $e =
12\,\si{\milli\meter}$, dropping a gross head $H = 90\,\si{\meter}$ to
a turbine. The penstock steel has Young's modulus $E =
200\,\text{GPa}$; the water has density $\rho =
1000\,\si{\kilogram\per\meter\cubed}$ and bulk modulus $K =
2.2\,\text{GPa}$. At full load the turbine passes $Q =
3.4\,\si{\meter\cubed\per\second}$. The pipe is rated to a maximum
working pressure of $1.6\,\text{MPa}$. The question is whether a fast
emergency closure of the turbine inlet valve can be survived, and if
not, what closure schedule or mitigation keeps the surge inside the
rating.

\paragraph{Step 1: the steady state.} The penstock area is $A =
\tfrac{\pi}{4}D^2 = \tfrac{\pi}{4}(1.2)^2 = 1.131\,\si{\meter}^2$, so
the steady flow speed is
\[
u_0 = \frac{Q}{A} = \frac{3.4}{1.131} = 3.0\,\si{\meter\per\second}.
\]
The static pressure at the foot, from Bernoulli on the free-surface
streamline (reservoir at rest, the foot moving at $u_0$, both at the
same kinetic correction), is dominated by the gross head:
$p_\text{foot} \approx \rho g H = 1000\times 9.81\times 90 =
0.88\,\text{MPa}$, comfortably inside the $1.6\,\text{MPa}$ rating.
Steady operation is not the concern; the transient is.

\paragraph{Step 2: the wave speed with pipe elasticity.} The pressure
wave does not travel at the sonic speed of the water alone, because
the pipe wall stretches as the pressure rises and stores some of the
fluid that the compression would otherwise have to absorb. The
Korteweg formula corrects the water sonic speed
$a_w = \sqrt{K/\rho}$ for the wall compliance of a thin-walled pipe:
\[
a = \sqrt{\frac{K/\rho}{1 + (K D)/(E e)}}.
\]
The water-alone speed is $a_w = \sqrt{2.2\times 10^9/1000} =
1483\,\si{\meter\per\second}$. The compliance term is
\[
\frac{K D}{E e} = \frac{(2.2\times 10^9)(1.2)}
{(200\times 10^9)(0.012)} = \frac{2.64\times 10^9}{2.4\times 10^9}
= 1.10,
\]
so
\[
a = \frac{1483}{\sqrt{1 + 1.10}} = \frac{1483}{1.449}
= 1024\,\si{\meter\per\second}.
\]
The elastic pipe slows the wave by nearly a third relative to rigid
confinement. We round to $a = 1000\,\si{\meter\per\second}$ for the
estimates that follow.

\paragraph{Step 3: the Joukowsky surge for instantaneous closure.}
Stopping the column instantaneously converts its momentum into a
pressure spike given by the Joukowsky relation $\Delta p = \rho a
\Delta u$, with $\Delta u = u_0 = 3.0\,\si{\meter\per\second}$:
\[
\Delta p = 1000\times 1000\times 3.0 = 3.0\,\text{MPa}.
\]
Added to the steady $0.88\,\text{MPa}$, the peak pressure at the
valve reaches $3.9\,\text{MPa}$, well above the $1.6\,\text{MPa}$
rating. An instantaneous (or sufficiently fast) closure ruptures the
penstock. The verdict at this step is a fail, and the rest of the
case study is the mitigation.

\paragraph{Step 4: the closure-time threshold.} The full Joukowsky
surge develops only when the valve closes before the relief wave can
return from the reservoir. The wave travels to the upper end and back
in the round-trip time
\[
t_r = \frac{2L}{a} = \frac{2\times 600}{1000} = 1.2\,\si{\second}.
\]
A closure faster than $1.2\,\si{\second}$ is hydraulically
``instantaneous'' and produces the full $3.0\,\text{MPa}$ surge. A
closure slower than $t_r$ lets successive reflections cancel part of
the rise; the arithmetic-surge approximation estimates the reduced
peak as
\[
\Delta p_\text{slow} \approx \rho a u_0 \,\frac{2L/a}{t_c}
= \frac{\rho L u_0}{t_c}\cdot 2,
\]
inversely proportional to the closure time
(Figure~\ref{fig:vol03ch11:waterhammer-surge}). To hold the surge to
a margin below the rating, say to $\Delta p_\text{allow} =
0.7\,\text{MPa}$ on top of the steady $0.88\,\text{MPa}$ (keeping the
peak at $1.58\,\text{MPa}$, just inside the rating), the closure must
satisfy
\[
t_c \ge \frac{2L u_0}{\Delta p_\text{allow}/\rho}
= \frac{2\times 600\times 3.0}{0.7\times 10^6/1000}
= \frac{3600}{700} = 5.1\,\si{\second}.
\]
A closure spread over at least about five seconds keeps the surge
inside the rating; the turbine governor and the valve actuator are
specified to that minimum.

\input{volumes/03-force/ch11-fluid-dynamics-inviscid/figures/fig-waterhammer-surge.tex}

\paragraph{Step 5: the surge tank as the alternative mitigation.} A
slow valve is not always acceptable; an emergency trip may need to
stop the flow faster than five seconds. The standard alternative is a
surge tank near the upper end of the penstock, a vertical riser open
at the top that gives the decelerating column a free surface to push
into. When the valve shuts, the column upstream of the tank carries
on flowing into the tank, its level rising until the reservoir head
arrests it; the surge then oscillates between tank and reservoir at a
period of minutes, slow enough that the pressure rise is the modest
hydrostatic level change rather than the acoustic Joukowsky spike. The
peak level rise above the reservoir, for a frictionless surge,
follows from energy conservation of the oscillating column,
\[
z_\text{max} = u_0 \sqrt{\frac{L A}{g A_t}},
\]
where $A_t$ is the surge-tank cross-section. Sizing the tank at
$A_t = 7\,\si{\meter}^2$,
\[
z_\text{max} = 3.0\sqrt{\frac{600\times 1.131}{9.81\times 7}}
= 3.0\sqrt{\frac{678.6}{68.7}} = 3.0\times 3.14
= 9.4\,\si{\meter},
\]
a pressure rise of only $\rho g z_\text{max} = 1000\times 9.81\times
9.4 = 0.092\,\text{MPa}$ on the upstream side. The penstock below the
tank still feels the Joukowsky surge of the short reach between tank
and valve, but that reach is a small fraction of $L$, so its
round-trip time is short and even a moderately fast valve stays inside
the rating. The surge tank decouples the long upstream column from the
fast valve action; it is why almost every long high-head penstock has
one.

\paragraph{Verdict.} The bare penstock with a fast valve fails:
$3.9\,\text{MPa}$ against a $1.6\,\text{MPa}$ rating. Two mitigations
each bring it inside the rating: a governed closure of at least about
five seconds, or a surge tank that confines the fast-valve surge to a
short lower reach while limiting the upstream rise to under a tenth of
a megapascal. The decision between them is operational, the surge tank
costs a structure but permits fast emergency trips; the slow closure
costs nothing but constrains the protection scheme. The inviscid
relations of this chapter, Bernoulli for the steady head, the
Joukowsky relation for the surge, the elastic wave speed for the
timing, plus the elementary energy balance of the oscillating column,
together carry the design from a layout to a defensible verdict. None
of the analysis needed viscosity; the losses that viscosity adds
would only damp the surge and make the verdict conservative.

\subsection{The diagnostic principle for inviscid analysis}

The chapter's diagnostic question, applied to any flow under
inviscid analysis, is: do the five Bernoulli assumptions hold along
the streamline of interest, between the two points where the
equation is being applied?

\begin{itemize}[leftmargin=*]
  \item Is the flow steady? Check for time-varying boundary
    conditions, pulsations, and starting transients.
  \item Is the flow incompressible? Check the Mach number; if
    $M > 0.3$ at any point on the streamline, use the compressible
    forms of chapter 13.
  \item Is the flow inviscid along the streamline? Check whether
    the streamline passes through a boundary layer or a separated
    region; if so, Bernoulli does not apply across the viscous
    region.
  \item Is the integration along a streamline? Check whether the
    two endpoints are connected by a continuous streamline through
    a non-recirculating region.
  \item Is the flow free of heat addition, shaft work, and
    separation between the endpoints? A compressor, a turbine, a
    heater, or a separated wake between the endpoints invalidates
    the relation.
\end{itemize}

A streamline that passes all five checks is fair game for
Bernoulli; a streamline that fails any one is not. The discipline
is to identify which assumption is violated and to use the
appropriate corrective tool: the momentum equation for energy-loss
regions, the unsteady Bernoulli for time-varying flows, the
compressible analysis for high Mach, the boundary-layer equations
for wall regions, the energy equation with work and heat terms for
turbomachinery.

\begin{principle}[Bernoulli applies under five assumptions, all checked]
Bernoulli's equation $\tfrac{1}{2}\rho u^2 + p + \rho g z = p_0$
holds along a streamline of a steady, incompressible, inviscid,
non-heat-receiving, non-work-receiving flow. A calculation that
applies Bernoulli to a streamline that violates any of the five
assumptions produces a number that looks plausible and is wrong.
The diagnostic discipline is to check the assumptions before
writing the equation, and to use the momentum, energy, or
boundary-layer equations when one or more assumptions fails.
\end{principle}

Two named cases connect this chapter's failure mode to the wider
volume. The aeroelastic flutter of the original Tacoma Narrows
suspension bridge in 1940
\cite{acc:tacoma-narrows-1941}, treated as a structural-dynamics
failure in chapter 6 of this volume and as a bluff-body
aerodynamic case in chapter 13, illustrates that an inviscid
pressure analysis of a body in steady wind cannot predict the
self-excited aerodynamic moment that drove the deck to oscillate;
the unsteady, separating flow over the deck section breaks four of
the five Bernoulli assumptions at once. The common-mode failure of
all three pitot probes on Air France Flight 447 on 1 June 2009
\cite{acc:bea-af447}, treated in chapter 12 of this volume,
illustrates that the Bernoulli-based velocity measurement of a
Pitot tube requires the freestream stagnation port to be free of
ice; ice crystals at altitude blocked the ports of all three
nominally independent probes, producing wrong airspeed indications
that the autopilot recognised as inconsistent and the flight crew
could not resolve. The Bernoulli physics was not at fault; the
boundary condition on the measurement (an unobstructed stagnation
port) was. Sensor independence in design did not deliver sensor
independence in failure, and the case has become a canonical
lesson in the limits of measurement under shared environmental
cause.

\section{Project}\pathtag{core}

\begin{project}[Build and calibrate a Pitot tube; measure household fan flow]
The reader builds a simple Pitot-static probe from common
materials, calibrates it against a known reference, and uses it to
map the velocity field downstream of a household fan. The
deliverable is a twenty- to thirty-page engineering report.

\textbf{Track}. Physical build plus measurement plus analysis.
\textbf{Hazard class}: low.

\textbf{Construction}. The probe consists of a forward-facing tube
(stagnation port) and a side-vented tube (static port), each
connected to one side of a differential pressure sensor or a
U-tube manometer. Suitable materials: stainless-steel hypodermic
tubing of two diameters (the inner stagnation tube nested inside an
outer static tube with side holes), or two separate rigid plastic
tubes mounted parallel with appropriate end geometries. The static
port should be at least four tube diameters back from the leading
edge to be free of the stagnation-region pressure rise.

\textbf{Calibration}. The reader calibrates the probe against a
known flow speed. Options:

\begin{itemize}[leftmargin=*]
  \item A pendulum-style anemometer (a light flap on a hinge,
    calibrated against the deflection at a known wind), useful for
    speeds above $1\,\si{\meter\per\second}$.
  \item A free-fall calibration: the probe is mounted on a moving
    arm or a falling carriage of known velocity, the differential
    pressure recorded at the known speed.
  \item A reference from a fan's manufacturer specification at a
    known rotational speed (typically accurate to within
    $\pm 20\,\%$, sufficient for the calibration of an amateur
    probe).
\end{itemize}

The calibration produces a relation $u = K\sqrt{2 \Delta p/\rho}$,
where $K$ is a probe constant close to $1.0$ for well-built
probes. The reduction script \path{code/pitot_calibration.py} fits
$K$ from reference points by least squares, applies it to a radial
survey, and integrates the axisymmetric profile to a volume flow
rate; the worked survey it writes to \path{data/pitot_fan_survey.csv}
fits $K = 0.997$ and returns about $700\,\si{\meter\cubed}$ per hour,
a template the reader adapts to the measured fan data.

\textbf{Mapping}. The reader maps the velocity field downstream of
a small household fan (axial blade type, output diameter $\sim
100$ to $200\,\si{\milli\meter}$). The probe is positioned at a
grid of points downstream of the fan: at least three streamwise
distances (one fan-diameter downstream, three, ten) and at least
five radial positions per streamwise distance. The reader records
the dynamic pressure at each grid point and converts to local
velocity using the calibrated probe constant.

\textbf{Analysis}. The reader produces:

\begin{itemize}[leftmargin=*]
  \item A plot of velocity profile $u(r)$ at each streamwise
    distance. The expected behaviour: a peaked profile near the
    fan with a velocity defect along the axis (the hub region),
    transitioning to a more uniform profile downstream as the wake
    spreads.
  \item An estimate of the total volume flow rate $Q = \int u \,dA$
    by numerical integration of the velocity profile, compared to
    the fan's nameplate rating.
  \item A discussion of the discrepancies between the inviscid
    Bernoulli expectation (uniform downstream velocity, no
    spreading) and the measured profile, naming the viscous and
    rotational effects that the inviscid analysis omits.
\end{itemize}

\textbf{Reflection ($1{,}500$ to $3{,}000$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
  \item The probe's accuracy; the source of the largest
    measurement uncertainty (probe alignment, tube response time,
    sensor calibration, or otherwise).
  \item The departure of the measured flow from the inviscid
    Bernoulli prediction, named in physical terms (boundary-layer
    growth, wake spreading, hub recirculation).
  \item One design improvement to the probe and one design
    improvement to the calibration that would tighten the
    measurement, with a brief analysis of why each would help.
\end{itemize}

\textbf{Deliverable}. The probe (photographed and dimensioned),
the calibration data, the mapping data, the analysis, and the
reflection.

\textbf{Time}. Two to four hours of probe construction, two to
four hours of calibration, three to six hours of mapping, four to
eight hours of write-up.
\end{project}

\section{Exercises}

The exercises mix calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A horizontal pipe carries water at $1.5\,\si{\meter\per\second}$ in
a section of $100\,\si{\milli\meter}$ diameter. The pipe contracts
smoothly to $50\,\si{\milli\meter}$ diameter. Find the flow speed
and the pressure drop in the narrower section, assuming inviscid
flow.

\paragraph{Solution sketch.} Continuity sets the speed: the area
ratio is $(100/50)^2 = 4$, so $u_2 = 4 u_1 = 4\times 1.5 =
6\,\si{\meter\per\second}$. Bernoulli on a horizontal streamline
($z_1 = z_2$) gives the pressure drop $p_1 - p_2 = \tfrac12\rho(u_2^2
- u_1^2) = 0.5\times 1000\times(36 - 2.25) = 1.69\times
10^4\,\text{Pa} \approx 16.9\,\text{kPa}$. The narrow section runs
$16.9\,\text{kPa}$ below the wide section.
\end{exercise}

\begin{exercise}
A Pitot tube on an aircraft reads a stagnation-static differential
pressure of $5\,\text{kPa}$. The air density at the flight
altitude is $1.0\,\si{\kilogram\per\meter\cubed}$. Find the
airspeed.

\paragraph{Solution sketch.} The Pitot relation gives $u =
\sqrt{2(p_0 - p)/\rho} = \sqrt{2\times 5000/1.0} = \sqrt{10\,000} =
100\,\si{\meter\per\second}$. The reading is an indicated airspeed
based on the assumed density; at the lower true density of altitude
the true airspeed is higher than this indicated value, which is why
the density used must be the local one.
\end{exercise}

\begin{exercise}
A water tank has a small hole $2\,\si{\meter}$ below the free
surface. Find the speed of the water emerging from the hole and
the volume flow rate through a hole of $10\,\si{\milli\meter}$
diameter, assuming a discharge coefficient of $0.62$.

\paragraph{Solution sketch.} Torricelli gives the ideal jet speed
$u = \sqrt{2gh} = \sqrt{2\times 9.81\times 2} =
6.26\,\si{\meter\per\second}$. The hole area is $A = \pi(0.005)^2 =
7.85\times 10^{-5}\,\si{\meter}^2$. The actual discharge applies the
coefficient, $Q = C_d A u = 0.62\times 7.85\times 10^{-5}\times 6.26
= 3.05\times 10^{-4}\,\si{\meter\cubed\per\second} \approx
0.30\,\text{L/s}$. The coefficient absorbs the jet contraction (the
vena contracta) and the small viscous loss.
\end{exercise}

\begin{exercise}
A Venturi meter in a horizontal pipe has $100\,\si{\milli\meter}$
inlet diameter and $40\,\si{\milli\meter}$ throat diameter, with
$C_d = 0.97$. The measured static pressure difference is
$30\,\text{kPa}$, the fluid is water. Find the volume flow rate.

\paragraph{Solution sketch.} The areas are $A_1 = \pi(0.05)^2 =
7.85\times 10^{-3}\,\si{\meter}^2$ and $A_2 = \pi(0.02)^2 = 1.257
\times 10^{-3}\,\si{\meter}^2$, with area ratio $A_1/A_2 = 6.25$. The
inlet speed is $u_1 = \sqrt{2(p_1 - p_2)/[\rho((A_1/A_2)^2 - 1)]} =
\sqrt{2\times 30\,000/[1000\times(39.06 - 1)]} = \sqrt{60\,000/38\,060}
= 1.256\,\si{\meter\per\second}$. The ideal flow rate is $A_1 u_1 =
9.86\,\text{L/s}$; applying $C_d = 0.97$ gives $Q = 9.57\,\text{L/s}$,
matching \path{code/venturi_flow.py}.
\end{exercise}

\begin{exercise}
Air flows steadily over a flat plate at $50\,\si{\meter\per\second}$
in standard sea-level air ($\rho = 1.225\,\si{\kilogram\per\meter
\cubed}$). Find the dynamic pressure and the stagnation pressure
just above the plate, given an ambient static pressure of
$101.3\,\text{kPa}$.

\paragraph{Solution sketch.} The dynamic pressure is $q = \tfrac12
\rho U^2 = 0.5\times 1.225\times 50^2 = 1531\,\text{Pa} \approx
1.53\,\text{kPa}$. The stagnation pressure is the static plus the
dynamic, $p_0 = p + q = 101.3 + 1.53 = 102.8\,\text{kPa}$. At Mach
$50/340 \approx 0.15$ the incompressible form is accurate to under
$1\,\%$.
\end{exercise}

\begin{exercise}
A symmetric thin airfoil at $4^{\circ}$ angle of attack in a
$60\,\si{\meter\per\second}$ airstream ($\rho =
1.0\,\si{\kilogram\per\meter\cubed}$) has chord $1.5\,\si{\meter}$.
Compute the lift coefficient, the bound circulation per unit span,
and the lift per unit span.

\paragraph{Solution sketch.} Convert the angle: $\alpha = 4^\circ =
0.0698\,\text{rad}$. The lift coefficient is $C_L = 2\pi\alpha =
0.439$. The dynamic pressure is $q = \tfrac12\rho U^2 = 0.5\times 1.0
\times 60^2 = 1800\,\text{Pa}$, so the lift per unit span is $L' =
C_L\,q\,c = 0.439\times 1800\times 1.5 = 1185\,\si{\newton\per
\meter}$. The bound circulation follows from Kutta-Joukowski, $\Gamma
= L'/(\rho U) = 1185/(1.0\times 60) = 19.7\,\si{\meter\squared\per
\second}$, equal to $\pi c U\alpha = \pi\times 1.5\times 60\times
0.0698$ as a check.
\end{exercise}

\begin{exercise}
A water main of $300\,\si{\milli\meter}$ diameter carries water at
$2\,\si{\meter\per\second}$. The main rises $15\,\si{\meter}$
between two measurement points. Find the static pressure at the
upper point if the lower point has $400\,\text{kPa}$ static
pressure, assuming inviscid steady flow.

\paragraph{Solution sketch.} The pipe has constant diameter, so
continuity gives $u_1 = u_2 = 2\,\si{\meter\per\second}$ and the
velocity heads cancel. Bernoulli then reduces to the hydrostatic
balance $p_2 = p_1 - \rho g(z_2 - z_1) = 400\,000 - 1000\times 9.81
\times 15 = 400\,000 - 147\,150 = 253\,\text{kPa}$. The static
pressure drops by the elevation head $\rho g\,\Delta z$ alone; raising
the water $15\,\si{\meter}$ costs about $147\,\text{kPa}$.
\end{exercise}

\begin{exercise}
A circular cylinder of radius $50\,\si{\milli\meter}$ is in a
uniform freestream of $10\,\si{\meter\per\second}$ in water
($\rho = 1000\,\si{\kilogram\per\meter\cubed}$). With no
circulation, find the velocity and pressure coefficient at the top
of the cylinder. State why the integrated pressure gives zero
drag.

\paragraph{Solution sketch.} At the top, $\theta = 90^\circ$, the
surface speed is $u_\theta = -2U\sin\theta = -2\times 10\times 1 =
-20\,\si{\meter\per\second}$ (magnitude $20\,\si{\meter\per\second}$,
twice the freestream). The pressure coefficient is $C_p = 1 -
4\sin^2\theta = 1 - 4 = -3$, so the gauge pressure is $p - p_\infty =
C_p\,\tfrac12\rho U^2 = -3\times 0.5\times 1000\times 100 = -1.5\times
10^5\,\text{Pa} = -150\,\text{kPa}$. The integrated pressure gives
zero drag because $C_p(\theta) = 1 - 4\sin^2\theta$ is symmetric
front to back: the pressure on the upstream face is matched by an
equal pressure on the downstream face, so the streamwise components
cancel (d'Alembert's paradox). The radius value is not needed for
$C_p$.
\end{exercise}

\begin{exercise}
A sharp-crested rectangular weir of width $2\,\si{\meter}$ has a
head of $0.30\,\si{\meter}$ above the crest. Using $C_d = 0.62$,
find the discharge over the weir.

\paragraph{Solution sketch.} The rectangular-weir formula gives $Q =
C_d\,\tfrac23 b\sqrt{2g}\,H^{3/2} = 0.62\times 0.667\times 2\times
\sqrt{19.62}\times 0.30^{1.5}$. Evaluating, $\sqrt{19.62} = 4.43$ and
$0.30^{1.5} = 0.164$, so $Q = 0.62\times 0.667\times 2\times 4.43
\times 0.164 = 0.60\,\si{\meter\cubed\per\second}$. The $H^{3/2}$
scaling means a $10\,\%$ rise in head raises the discharge by about
$15\,\%$, which is what makes a weir a sensitive flow gauge.
\end{exercise}

\begin{exercise}
A garden hose ($20\,\si{\milli\meter}$ diameter) carries water at
$0.5\,\si{\meter\per\second}$. The end is partially covered so the
exit area is reduced by a factor of four. Find the exit jet speed,
the dynamic pressure at the exit, and the supply pressure required
upstream, assuming inviscid flow and atmospheric exit pressure.

\paragraph{Solution sketch.} Continuity with a fourfold area
reduction gives an exit speed $u_e = 4\times 0.5 =
2\,\si{\meter\per\second}$. The exit dynamic pressure is $q_e =
\tfrac12\rho u_e^2 = 0.5\times 1000\times 4 = 2000\,\text{Pa} =
2\,\text{kPa}$. Bernoulli from the hose interior (speed
$0.5\,\si{\meter\per\second}$) to the atmospheric exit gives the
required gauge supply pressure $p_s - p_\text{atm} = \tfrac12\rho
(u_e^2 - u_1^2) = 0.5\times 1000\times(4 - 0.25) =
1875\,\text{Pa}\approx 1.9\,\text{kPa}$ above atmospheric. The supply
pressure exceeds the exit dynamic pressure only by the small inlet
kinetic term.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the differential continuity equation $\partial \rho/
\partial t + \nabla\cdot(\rho\vec{u}) = 0$ from the integral form,
stating the assumptions on the control volume.

\paragraph{Solution sketch.} Start from $\frac{d}{dt}\int_\mathcal{V}
\rho\,dV + \oint_{\partial\mathcal{V}}\rho\vec{u}\cdot\hat{n}\,dA =
0$. For a fixed (time-independent) control volume the time derivative
passes inside as a partial: $\int_\mathcal{V}\partial\rho/\partial
t\,dV$. The divergence theorem turns the surface integral into
$\int_\mathcal{V}\nabla\cdot(\rho\vec{u})\,dV$, which requires
$\rho\vec{u}$ to be continuously differentiable in $\mathcal{V}$.
Combining, $\int_\mathcal{V}[\partial\rho/\partial t +
\nabla\cdot(\rho\vec{u})]\,dV = 0$. Since the control volume is
arbitrary, the integrand must vanish pointwise, giving the
differential form. The assumptions are a fixed control volume and a
smooth flow field (no shocks or interfaces inside $\mathcal{V}$).
\end{exercise}

\begin{exercise}
Derive Bernoulli's equation $\tfrac{1}{2}\rho u^2 + p + \rho g z =
p_0$ from the steady incompressible Euler equation by integration
along a streamline. State each assumption as it is invoked.

\paragraph{Solution sketch.} Start from steady Euler, $(\vec{u}\cdot
\nabla)\vec{u} = -\tfrac1\rho\nabla p + \vec{g}$ (steady assumption
drops $\partial\vec{u}/\partial t$). Dot with the unit tangent
$\hat{s}$ and use $(\vec{u}\cdot\nabla)\vec{u}\cdot\hat{s} =
u\,\partial u/\partial s = \tfrac12\partial(u^2)/\partial s$. The
pressure term becomes $-\tfrac1\rho\partial p/\partial s$ and gravity
$-g\,\partial z/\partial s$. With $\rho$ constant (incompressible),
multiply by $\rho$ and integrate along $s$ from point 1 to point 2:
$\int(\tfrac\rho2\partial(u^2)/\partial s + \partial p/\partial s +
\rho g\,\partial z/\partial s)\,ds = 0$, yielding $\tfrac12\rho u^2 +
p + \rho g z = \text{const}$. The inviscid assumption entered when
Euler omitted viscous stress; the along-a-streamline integration
fixes the constant only on that streamline.
\end{exercise}

\begin{exercise}
Show that the velocity potential $\phi$ for an irrotational
incompressible flow satisfies Laplace's equation $\nabla^2 \phi =
0$.

\paragraph{Solution sketch.} Irrotationality means $\nabla\times
\vec{u} = 0$, which (in a simply connected region) lets the velocity
be written as a gradient, $\vec{u} = \nabla\phi$; this guarantees
$\nabla\times\nabla\phi = 0$ identically. Incompressibility gives
$\nabla\cdot\vec{u} = 0$. Substituting the first into the second,
$\nabla\cdot(\nabla\phi) = \nabla^2\phi = 0$. The two physical
conditions, no rotation and no volume change, combine into Laplace's
equation; its harmonic solutions are the same class that governs
electrostatic potential and steady heat conduction.
\end{exercise}

\begin{exercise}
Derive the Kutta-Joukowski theorem $L' = \rho U_\infty \Gamma$ by
integrating the pressure around a closed body in a uniform inviscid
flow with bound circulation $\Gamma$.

\paragraph{Solution sketch.} The compact route uses the complex
potential. Write the complex velocity far from the body as a Laurent
series $w(z) = dF/dz = U_\infty + \Gamma/(2\pi i z) + \cdots$. The
Blasius force theorem gives the force per unit span as $F_x - iF_y =
\tfrac{i\rho}{2}\oint (w)^2\,dz$. Substituting the series and keeping
the residue (the $1/z$ term survives the contour integral), the
integral picks up $2\pi i$ times the coefficient of $1/z$ in $w^2$,
which is $2U_\infty\,\Gamma/(2\pi i)$. The algebra yields $F_x = 0$
(no drag, d'Alembert) and $F_y = \rho U_\infty\Gamma$, the
Kutta-Joukowski lift. The real-variable route integrates the
Bernoulli pressure $p = p_0 - \tfrac12\rho|\vec{u}|^2$ over the
surface with the asymmetric speed produced by the added circulation;
only the cross term between freestream and circulation survives, again
giving $\rho U_\infty\Gamma$.
\end{exercise}

\begin{exercise}
Show that, for a converging-diverging nozzle in inviscid
incompressible flow, the pressure recovery is complete if the exit
area equals the inlet area. State why real diffusers do not
recover the full dynamic pressure.

\paragraph{Solution sketch.} Continuity gives $u_\text{exit} =
u_\text{inlet}(A_\text{inlet}/A_\text{exit})$; with equal areas the
exit speed equals the inlet speed. Bernoulli on the centreline (no
elevation change) then gives $p_\text{exit} = p_\text{inlet} +
\tfrac12\rho(u_\text{inlet}^2 - u_\text{exit}^2) = p_\text{inlet}$:
the pressure returns exactly. Real diffusers fall short because the
diverging section runs against an adverse pressure gradient, the
boundary layer thickens and separates, and the separated region
dissipates mechanical energy as turbulence; the lost energy is not
available to recover pressure, so a good diffuser recovers only
$70$ to $90\,\%$ of the dynamic pressure.
\end{exercise}

\begin{exercise}
Derive the unsteady Bernoulli equation $\partial \phi/\partial t +
\tfrac{1}{2}u^2 + p/\rho + gz = f(t)$ for irrotational
incompressible unsteady flow. Identify the term that vanishes
under the steady assumption.

\paragraph{Solution sketch.} For irrotational flow $\vec{u} =
\nabla\phi$. Substitute into the full Euler equation and use the
vector identity $(\vec{u}\cdot\nabla)\vec{u} = \nabla(\tfrac12 u^2) -
\vec{u}\times(\nabla\times\vec{u})$; the last term vanishes because
the flow is irrotational. The local-acceleration term is
$\partial\vec{u}/\partial t = \nabla(\partial\phi/\partial t)$. Euler
then reads $\nabla[\partial\phi/\partial t + \tfrac12 u^2 + p/\rho +
gz] = 0$, so the bracket is a function of time alone, $f(t)$. The
term that vanishes under the steady assumption is $\partial\phi/
\partial t$; dropping it recovers the ordinary Bernoulli equation
with $f(t)$ a constant.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the lift produced by a typical paper airplane at
release. State assumptions about the airplane's wing area, angle
of attack, release speed, and air density. Compare to the
airplane's weight.

\paragraph{Reference estimate.} Take wing area $A \approx
0.03\,\si{\meter}^2$, release speed $U \approx 6\,\si{\meter\per
\second}$, angle of attack $\alpha \approx 5^\circ$, air density
$1.2\,\si{\kilogram\per\meter\cubed}$, and (for a low-aspect-ratio
flat plate, below the thin-airfoil slope) $C_L \approx 0.3$. Then $L =
C_L\,\tfrac12\rho U^2 A = 0.3\times 0.5\times 1.2\times 36\times 0.03
\approx 0.19\,\si{\newton}$. A sheet of paper massing about
$5\,\si{\gram}$ weighs $0.05\,\si{\newton}$, so the lift is a few
times the weight, consistent with a plane that glides rather than
drops. The estimate is good to a factor of two or three; the largest
uncertainty is $C_L$ at low aspect ratio.
\end{exercise}

\begin{exercise}
Estimate the maximum flow speed downstream of a household sink tap
when fully opened. State assumptions about the supply pressure,
the tap geometry, and the length of any vertical drop from supply
to spout.

\paragraph{Reference estimate.} Household mains pressure is about
$300\,\text{kPa}$ gauge ($3\,\text{bar}$). Treating the tap as an
opening to atmosphere and applying Bernoulli from the pressurised
supply to the open spout, the ideal jet speed is $u = \sqrt{2\Delta
p/\rho} = \sqrt{2\times 300\,000/1000} = \sqrt{600} \approx
24\,\si{\meter\per\second}$. Real taps run far slower because the
valve, aerator, and pipe friction drop most of the head; a measured
spout speed of $2$ to $4\,\si{\meter\per\second}$ is typical, so the
inviscid Bernoulli value is an upper bound roughly $6$ to $10$ times
the real speed. The gap is the chapter's standing warning: Bernoulli
sets the ceiling, not the operating point, when losses dominate.
\end{exercise}

\begin{exercise}
Estimate the volume flow rate through a leaking garden hose with a
pinhole at the supply (mains) end. State assumptions about the
supply pressure and the pinhole diameter; bracket the answer to
within a factor of three.

\paragraph{Reference estimate.} Take supply pressure $300\,\text{kPa}$
gauge and a pinhole of diameter $1\,\si{\milli\meter}$, area $A =
\pi(0.0005)^2 = 7.85\times 10^{-7}\,\si{\meter}^2$. The Torricelli
speed is $u = \sqrt{2\Delta p/\rho} = \sqrt{600} \approx
24\,\si{\meter\per\second}$. With a discharge coefficient $C_d
\approx 0.6$ for a sharp hole, $Q = C_d A u = 0.6\times 7.85\times
10^{-7}\times 24 \approx 1.1\times 10^{-5}\,\si{\meter\cubed\per
\second} \approx 0.7\,\text{L/min}$. Bracketing the pinhole between
$0.5$ and $2\,\si{\milli\meter}$ (a factor of four in area) spans
roughly $0.2$ to $3\,\text{L/min}$, inside the requested factor of
three about a central estimate near $0.7\,\text{L/min}$.
\end{exercise}

\begin{exercise}
Estimate the wind speed that would cause a sheet of office paper
to lift off a flat desk. State the paper's mass per unit area and
the relevant pressure-coefficient differential between upper and
lower surfaces.

\paragraph{Reference estimate.} Office paper masses about
$80\,\si{\gram\per\meter\squared}$, so its weight per unit area is
$0.08\times 9.81 \approx 0.78\,\si{\pascal}$. Lift requires a
pressure differential of this order across the sheet. A wind blowing
over the top creates a suction; take a pressure-coefficient
differential $\Delta C_p \approx 0.5$. Then $\tfrac12\rho U^2\,\Delta
C_p = 0.78$ gives $U^2 = 2\times 0.78/(1.2\times 0.5) = 2.6$, so $U
\approx 1.6\,\si{\meter\per\second}$. A gentle draught of a couple of
metres per second lifts a desk sheet, matching everyday experience;
the dominant uncertainty is $\Delta C_p$, which depends on how the
sheet sits and curls.
\end{exercise}

\begin{exercise}
Estimate the time required to drain a swimming pool of $25\,\si{
\meter} \times 10\,\si{\meter}$ surface area and $2\,\si{\meter}$
depth through a $100\,\si{\milli\meter}$ outlet near the bottom.
State assumptions about the discharge coefficient and the
behaviour as the surface drops to the outlet.

\paragraph{Reference estimate.} The closed-form drain time for a
constant-area surface is $t = (A_s/C_d A_o)\sqrt{2h_0/g}$. With $A_s =
250\,\si{\meter}^2$, $A_o = \pi(0.05)^2 = 7.85\times
10^{-3}\,\si{\meter}^2$, $C_d = 0.62$, and $h_0 = 2\,\si{\meter}$:
$\sqrt{2\times 2/9.81} = 0.639\,\si{\second}^{1/2}\cdot\si{\meter}^{1/2}$
factor, and $t = (250/(0.62\times 7.85\times 10^{-3}))\times 0.639
\approx 51\,400\times 0.639 \approx 3.3\times 10^4\,\si{\second}
\approx 9\,\text{h}$. The $\sqrt{h}$ dependence means the last shallow
layer drains slowly; once the surface nears the outlet the simple
Torricelli model breaks down (vortexing and air entrainment), so the
estimate is reliable for the bulk of the drain and optimistic for the
final centimetres. The script \path{code/tank_drain.py} confirms the
nine-hour figure.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a numerical solver for the two-dimensional Laplace
equation $\nabla^2 \phi = 0$ on a rectangular domain with mixed
Dirichlet and Neumann boundary conditions, using finite differences.
Verify on the flow around a circular cylinder by comparing the
computed surface pressure coefficient to the analytical $C_p =
1 - 4\sin^2\theta$.

\paragraph{Solution sketch.} Discretise the domain on a grid and
apply the five-point Laplacian; Gauss-Seidel or successive
over-relaxation iterates $\phi_{i,j} = \tfrac14(\phi_{i+1,j} +
\phi_{i-1,j} + \phi_{i,j+1} + \phi_{i,j-1})$ to convergence. Impose
$\phi = U_\infty x$ on the far field (Dirichlet) and zero normal
derivative on the cylinder surface (Neumann, no penetration). Recover
surface velocity by differencing $\phi$ tangentially, then $C_p = 1 -
(u_s/U_\infty)^2$. The computed distribution should reproduce the
peak $C_p = 1$ at the stagnation points and $C_p = -3$ at the top and
bottom; \path{code/cylinder_potential.py} gives the analytical target
to compare against. Convergence and grid-refinement checks are part
of the deliverable.
\end{exercise}

\begin{exercise}
Implement a point-vortex panel method for thin airfoils: discretise
the chord into panels, place a point vortex at the quarter-chord
of each panel, enforce the flow-tangency condition at the
three-quarter-chord of each panel, and solve the resulting linear
system. Verify on a flat plate by comparing the computed lift
slope to $C_L = 2\pi\alpha$.

\paragraph{Solution sketch.} Lay $N$ panels along the chord; place a
point vortex of unknown strength $\gamma_j$ at each panel's
quarter-chord and a collocation point at each three-quarter-chord.
The flow-tangency condition at each collocation point gives one
equation: the sum of induced normal velocities from all vortices plus
the freestream normal component $U_\infty\sin\alpha$ must vanish. This
is an $N\times N$ linear system $\mathbf{A}\boldsymbol{\gamma} =
\mathbf{b}$; solve it, sum the circulations $\Gamma = \sum\gamma_j$,
and form $C_L = \Gamma/(\tfrac12 U_\infty c)$. For a flat plate the
result converges to $C_L = 2\pi\alpha$ as $N$ grows; even $N = 1$ with
the quarter/three-quarter placement gives the exact slope, which is
the classical reason for that placement (the lumped-vortex theorem).
\end{exercise}

\begin{exercise}
Simulate the streaklines and pathlines of an unsteady flow field
$u_x = U_0 \cos(\omega t)$, $u_y = 0$, by integrating particle
trajectories. Show that the streamlines coincide with the pathlines
only when the flow is steady.

\paragraph{Solution sketch.} For the given field the pathline of a
particle from $(x_0,0)$ integrates to $x(t) = x_0 + (U_0/\omega)
\sin(\omega t)$, $y(t) = 0$: the particle oscillates back and forth
on a horizontal line. An instantaneous streamline at time $t_0$
follows $dx/u_x = dy/u_y$ with $u_y = 0$, so streamlines are also
horizontal lines, but their direction reverses with the sign of
$\cos(\omega t_0)$. A streakline of particles released continuously
from one point spreads along the same line with a phase-dependent
extent. With $u_y \ne 0$ added (as in the figure of this chapter) the
three curves separate visibly. The verification is to integrate
particle trajectories numerically and overlay them on the snapshot
streamlines: they coincide only when $\omega = 0$ (steady flow), and
the code should demonstrate the divergence for $\omega \ne 0$.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a Venturi flow meter to measure flow rates between $1$ and
$10\,\text{L/s}$ in a $50\,\si{\milli\meter}$ water pipe, with a
differential pressure transducer capable of resolving pressures
between $100\,\text{Pa}$ and $50\,\text{kPa}$. Specify the throat
diameter and check that the resulting differential pressures fall
within the transducer's range.

\paragraph{Model-answer sketch.} The pipe area is $A_1 = \pi(0.025)^2
= 1.96\times 10^{-3}\,\si{\meter}^2$. The differential pressure scales
as $\Delta p \approx \tfrac12\rho(Q/A_2)^2$ for a strong contraction.
Try a throat of $25\,\si{\milli\meter}$ ($A_2 = 4.91\times
10^{-4}\,\si{\meter}^2$, area ratio $4$): at $Q = 1\,\text{L/s}$, $u_2
= Q/A_2 = 2.04\,\si{\meter\per\second}$ and $\Delta p \approx
\tfrac12\rho(u_2^2 - u_1^2) \approx 2.0\,\text{kPa}$; at $Q =
10\,\text{L/s}$, $u_2 = 20.4\,\si{\meter\per\second}$ and $\Delta p
\approx 200\,\text{kPa}$, which exceeds the $50\,\text{kPa}$ ceiling.
Open the throat to $32\,\si{\milli\meter}$ (area ratio $\approx 2.4$):
the differential drops by roughly the square of the ratio change,
landing the $10\,\text{L/s}$ point near $45\,\text{kPa}$ and the
$1\,\text{L/s}$ point near $450\,\text{Pa}$, both inside the
transducer band. The design is iterative: pick the throat that keeps
$\Delta p$ inside the range across the full flow span, then verify
with \path{code/venturi_flow.py}.
\end{exercise}

\begin{exercise}
Design a sharp-crested rectangular weir to measure flows from a
small stream that varies between $0.05$ and $1\,\text{m}^3\text{
/s}$. Specify the crest width and the expected head range; check
that the discharge coefficient remains within typical limits over
the range.

\paragraph{Model-answer sketch.} Invert $Q = C_d\tfrac23 b\sqrt{2g}\,
H^{3/2}$ for $H$. Choose a crest width $b = 1\,\si{\meter}$. At $Q =
0.05\,\si{\meter\cubed\per\second}$: $H^{3/2} = Q/(C_d\tfrac23 b
\sqrt{2g}) = 0.05/(0.62\times 0.667\times 1\times 4.43) = 0.0273$, so
$H = 0.0273^{2/3} \approx 0.091\,\si{\meter}$. At $Q =
1\,\si{\meter\cubed\per\second}$: $H^{3/2} = 0.546$, $H \approx
0.66\,\si{\meter}$. The head ranges from about $90\,\si{\milli\meter}$
to $0.66\,\si{\meter}$, a comfortable measurement span. Check that the
low end keeps $H$ above roughly $30\,\si{\milli\meter}$ (below which
surface tension makes $C_d$ unreliable); $90\,\si{\milli\meter}$ is
safe. A wider crest lowers the heads and may push the low flow below
that limit, so $b = 1\,\si{\meter}$ is a reasonable choice.
\end{exercise}

\begin{exercise}
Design a household-scale wind-turbine blade of $2\,\si{\meter}$
diameter for operation in $5\,\si{\meter\per\second}$ winds at sea
level. Specify the blade's chord, twist, and angle of attack
distribution along the span, using thin-airfoil theory and the
inviscid Kutta-Joukowski relation. State which inviscid
assumptions break down and where, and estimate the resulting
viscous correction.

\paragraph{Model-answer sketch.} Pick a tip-speed ratio $\lambda
\approx 6$, so the tip speed is $\lambda V = 30\,\si{\meter\per
\second}$ and the local blade speed at radius $r$ is $\Omega r =
\lambda V r/R$. The local relative wind angle is $\phi = \arctan(V/
\Omega r)$, large near the root and small near the tip, which sets the
twist. Choose a working lift coefficient $C_L \approx 0.8$ at angle of
attack $\alpha \approx 5^\circ$, so the geometric pitch is $\phi -
\alpha$. The chord from blade-element-momentum theory scales roughly
as $c(r) \propto 1/(\lambda^2 r)$, giving a wide root tapering to a
narrow tip. The inviscid Kutta-Joukowski relation sets the lift per
element; the inviscid assumptions break down in the root region
(thick sections, low local Reynolds number, three-dimensional
flow) and near stall at low wind speeds. The viscous correction is a
drag-to-lift ratio of a few percent that lowers the power coefficient
below the Betz limit; expect a realised $C_P \approx 0.35$ to $0.45$.
\end{exercise}

\begin{exercise}
Design a converging-diverging diffuser to recover pressure from a
$30\,\si{\meter\per\second}$ air jet exiting a duct, with an exit
area double the inlet area. Specify the diffuser length and the
divergence angle; estimate the achievable pressure-recovery
coefficient.

\paragraph{Model-answer sketch.} Doubling the area halves the speed,
so the ideal pressure recovery is $\Delta p = \tfrac12\rho(u_1^2 -
u_2^2) = \tfrac12\rho u_1^2(1 - \tfrac14) = \tfrac34 q_1$, where $q_1
= \tfrac12\times 1.2\times 30^2 = 540\,\text{Pa}$, giving an ideal
$\Delta p = 405\,\text{Pa}$. To avoid separation keep the total
included divergence angle at or below about $7^\circ$ (each wall
$3.5^\circ$). For a circular duct, doubling the area multiplies the
diameter by $\sqrt2 = 1.41$; the length needed for a $3.5^\circ$
half-angle is $L = (d_2 - d_1)/(2\tan 3.5^\circ) \approx 0.41 d_1/
0.122 \approx 3.4\,d_1$. The real pressure-recovery coefficient
$C_p = \Delta p/q_1$ reaches about $0.6$ to $0.7$ of the ideal $0.75$
for a well-designed diffuser, so expect $C_p \approx 0.5$.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A homeowner reports that water pressure at the upper-floor shower
drops sharply when the kitchen tap is opened. Identify the
inviscid-fluid-mechanics calculation that quantifies the drop and
the design discipline that would have anticipated it. State the
diagnostic measurement.

\paragraph{Solution sketch.} Opening the kitchen tap adds flow to the
shared supply pipe; continuity raises the velocity in the common run,
and Bernoulli plus pipe-friction losses convert that higher velocity
into a static-pressure drop felt at the upstream branch to the shower.
The calculation is a head balance on the shared pipe: the head loss
scales with the square of the total flow, so the additional draw drops
the pressure available at the upper floor. The design discipline is
hydraulic network sizing, oversizing the trunk and branching close to
the source so that fixtures do not load a common small-diameter run.
The diagnostic measurement is a pressure gauge at the shower, read
with the kitchen tap shut and again with it open; the difference
localises the drop to the shared run.
\end{exercise}

\begin{exercise}
A Pitot tube on a small aircraft reads an airspeed $20\,\%$ lower
than the GPS-derived groundspeed. Possible causes: probe misalignment,
static-port obstruction, calibration drift, ice. Identify which of
these the diagnostic discipline would investigate first, and the
test that would distinguish among them.

\paragraph{Solution sketch.} Investigate the static-port obstruction
first: a partially blocked static port is the single fault that biases
the dynamic-pressure reading directly and is common and cheap to
check, whereas calibration drift is slow and probe misalignment
usually produces a smaller error. A $20\,\%$ low airspeed corresponds
to a dynamic-pressure error of about $36\,\%$ (since $u\propto
\sqrt{\Delta p}$), consistent with a static port reading too high.
The distinguishing test is to compare indicated airspeed against GPS
groundspeed on reciprocal headings in still air (averaging out wind),
and to inspect and clear the static port; if clearing the port
restores agreement, the cause is confirmed. Ice is distinguished by
its dependence on conditions (it appears in cloud or at low
temperature and clears with heat).
\end{exercise}

\begin{exercise}
A small wind turbine in steady $7\,\si{\meter\per\second}$ wind is
producing only $40\,\%$ of its nameplate power. Identify three
inviscid-flow-related explanations and the inspection that would
distinguish among them.

\paragraph{Solution sketch.} Three inviscid-flow explanations: (1) the
blades operate off their design tip-speed ratio, so the local angle of
attack is wrong and the bound circulation, hence lift, is low; (2) the
blade pitch or twist is misset, again moving the angle of attack off
the design point; (3) blade fouling or leading-edge damage alters the
effective camber and the Kutta condition, reducing circulation. The
distinguishing inspections: measure the rotor speed to compute the
actual tip-speed ratio and compare to design (explanation 1); check
the pitch setting against the design schedule (explanation 2); inspect
the leading edges and surfaces for fouling, ice, or erosion
(explanation 3). Power scales with the cube of the effective wind and
strongly with $C_P(\lambda)$, so a modest tip-speed-ratio error
explains a large power shortfall.
\end{exercise}

\begin{exercise}
A dam spillway has shown signs of cavitation damage on its concrete
face downstream of the crest. Identify the inviscid-pressure
calculation that would have predicted the local pressure drop and
the corresponding cavitation risk.

\paragraph{Solution sketch.} The calculation is the Bernoulli static
pressure along the spillway face: $u = \sqrt{2g(z_1 - z_2)}$ gives the
local speed and $p = p_\text{atm} - \tfrac12\rho u^2$ (plus any
$C_p$ deficit over surface irregularities) gives the local static
pressure. Where this drops below the vapour pressure $p_v \approx
2.3\,\text{kPa}$ absolute, the cavitation number $\sigma = (p -
p_v)/\tfrac12\rho u^2$ falls below inception and bubbles form, then
collapse downstream and pit the concrete (see
Figure~\ref{fig:vol03ch11:cavitation} and the
$30\,\si{\meter}$-spillway estimation). The discipline is to compute
$\sigma$ at every surface irregularity for the design discharge and,
where it is marginal, to add aeration; the diagnostic confirmation is
that damage clusters just downstream of joints and offsets where the
local pressure dips.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
The Air France Flight 447 accident report \cite{acc:bea-af447}
identifies pitot-icing-induced unreliable airspeed as a precipitating
condition. Argue, in $200$ to $300$ words, what the inviscid
Bernoulli physics of the Pitot tube does and does not guarantee
about the airspeed measurement, and what design discipline would
reduce the probability of a common-mode failure across three
nominally independent probes.

\paragraph{Model-answer sketch.} The Bernoulli physics guarantees
that, given an unobstructed stagnation port and an accurate static
pressure, the differential $p_0 - p$ equals the dynamic pressure
$\tfrac12\rho u^2$, so the airspeed is recoverable. It guarantees
nothing about the boundary condition on the measurement: it assumes
the stagnation port samples the true total pressure. Ice crystals
blocking the ports break that assumption, and the equation then reports
a wrong speed that looks valid. A strong answer notes that the three
probes were independent in wiring and electronics but shared one
environment, so an environmental cause (high-altitude ice) defeats the
redundancy: independence in design is not independence in failure
under a common cause. The design discipline is diversity, not just
redundancy: probes of differing heating power, geometry, or operating
principle, plus a model-based airspeed estimate from other sensors to
cross-check, so that no single environmental condition disables all
channels at once.
\end{exercise}

\begin{exercise}
A student applies Bernoulli's equation to a streamline that enters
a turbomachine inlet and exits the turbomachine outlet, getting an
inconsistent answer. Identify which of the five Bernoulli
assumptions is violated and the correct analytical tool.

\paragraph{Solution sketch.} The violated assumption is no shaft work
between the endpoints: a turbomachine adds or extracts mechanical
energy across the control volume, so the Bernoulli sum is not constant
along a streamline through the machine. The correct tool is the steady
energy equation with a work term, $\tfrac{p_1}\rho + \tfrac12 u_1^2 +
g z_1 + w_\text{shaft} = \tfrac{p_2}\rho + \tfrac12 u_2^2 + g z_2 +
\text{losses}$, where $w_\text{shaft}$ is the specific work of the
machine. Bernoulli is the special case with $w_\text{shaft} = 0$ and
zero losses; restoring the work term resolves the inconsistency.
\end{exercise}

\begin{exercise}
A garden-hose nozzle produces a jet whose computed Bernoulli speed
disagrees with the measured speed by a factor of two. Identify the
two most probable causes (one in the assumption set, one in the
measurement) and the diagnostic that would distinguish them.

\paragraph{Solution sketch.} Assumption-set cause: the nozzle is not
inviscid and not loss-free; pipe friction and the nozzle contraction
drop much of the supply head, so the real exit speed is well below the
Bernoulli value computed from the nominal supply pressure (the same
ceiling-not-operating-point gap as the sink-tap estimate). Measurement
cause: the speed was inferred from an indirect quantity (throw
distance, or a pressure read at the wrong point) rather than measured
directly. The diagnostic is a direct measurement: collect a known
volume over a timed interval to get $Q$, divide by the measured exit
area for the true mean speed, and measure the static pressure at the
nozzle inlet rather than at the supply; comparing the head actually
available at the nozzle to the nominal supply pressure shows whether
the loss (assumption) or the inference (measurement) is the larger
error.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A senior engineer argues that Bernoulli's equation is sufficient
for any subsonic-flow design problem because viscous corrections
are typically a few percent. Write a one-paragraph reply that
agrees where agreement is honest and disagrees where the
viscous-correction estimate fails.

\paragraph{Model-answer sketch.} Agree that for attached external
flow at moderate angle of attack, and for pressure recovery in
gently varying ducts, viscosity costs only a few percent and Bernoulli
gives the right leading-order answer; that is the working basis of
subsonic aerodynamics and flow metering. Disagree where the
viscous-correction estimate is not small: separated flow (the entire
drag of a bluff body, stall, diffusers past the divergence limit),
flows through boundary layers, hydraulic jumps, and any
energy-dissipating transition. In those cases the correction is not a
few percent but order one or larger, and Bernoulli is qualitatively
wrong; a senior engineer should treat the few-percent rule as valid
only after confirming the flow is attached and loss-free between the
points of interest.
\end{exercise}

\begin{exercise}
A junior engineer is taught the equal-transit-time explanation of
airfoil lift in an introductory physics course. Write a
one-paragraph response that identifies the explanation's errors
and states the correct physics in terms of bound circulation and
the Kutta condition.

\paragraph{Model-answer sketch.} The equal-transit-time story assumes
fluid particles split at the leading edge must rejoin at the trailing
edge; there is no such constraint, and on a real airfoil the upper
flow arrives well ahead of the lower flow. It also predicts no lift
for a flat plate or a symmetric airfoil, which plainly do lift at
angle of attack, and predicts negative lift for some cambered shapes
that lift. The correct physics: a real airfoil at angle of attack
develops a bound circulation $\Gamma$ fixed by the Kutta condition
(the flow leaves the sharp trailing edge smoothly). The circulation
adds to the freestream, producing a faster upper flow and a slower
lower flow; Bernoulli converts the speed difference into a pressure
difference, and the integrated difference is the lift $L' = \rho
U_\infty\Gamma$. The path length plays no role; the trailing-edge
condition and the resulting circulation do.
\end{exercise}

\begin{exercise}
A designer proposes to compute the drag of a road vehicle by
integrating the inviscid Bernoulli pressure distribution over the
vehicle's surface. Write a one-paragraph response covering why
this approach gives zero drag (d'Alembert's paradox), what
modification recovers a sensible drag estimate, and what physical
mechanism the inviscid analysis omits.

\paragraph{Model-answer sketch.} Integrating the inviscid Bernoulli
pressure over a closed body gives zero net streamwise force because
the surface pressure distribution is front-to-back symmetric: the
high pressure on the front is exactly recovered as high pressure on
the rear (d'Alembert's paradox, shown for the cylinder by $C_p = 1 -
4\sin^2\theta$). The modification that recovers a sensible estimate is
to include the wake: a real bluff body separates, the rear pressure
never recovers, and the front-to-rear pressure imbalance is the
pressure (form) drag, conventionally captured by a measured drag
coefficient $C_D$ so that $D = C_D\,\tfrac12\rho U^2 A$. The omitted
mechanism is viscosity, which creates the boundary layer, causes
separation, and (through skin friction) adds a second drag component;
the inviscid model has neither.
\end{exercise}

\begin{exercise}
Identify a fluid-flow situation in your immediate environment in
which Bernoulli's equation gives a useful first estimate, and one
in which it gives a misleading answer. State, in $200$ to $300$
words, the streamline you would draw in each case and the
assumption that fails in the misleading case.

\paragraph{Solution sketch.} A strong answer pairs a clean case with
a trap. Useful case: a partially covered garden hose or a tap aerator,
where a streamline from the supply to the open jet gives a sensible
exit-speed estimate because the flow is steady, nearly incompressible,
and short enough that losses are modest. Misleading case: a streamline
drawn through a hand-dryer or vacuum-cleaner nozzle and into the
turbulent room jet, or across a window in a gust, where the flow is
unsteady and separated; applying Bernoulli across the separated wake
violates the inviscid (and often the steady) assumption and gives a
wrong pressure. The discipline shown is to state the streamline
explicitly and check the five assumptions on it before trusting the
number.
\end{exercise}
