\chapter{Additive manufacturing}
\label{vol08:ch07}

\epigraph{The geometry that the process can make is not the geometry that the drawing shows.}{}

\chapmeta{Half-life: medium-life. Archetypes: scaling, optimisation, failure. Exercise target: 25. Project track: physical.}

\noindent
Additive manufacturing builds a part by adding material, layer by layer, from a digital description. The technology is neither a magic geometry shop nor a curiosity. It is a class of processes with characteristic capabilities, limits, defect modes, and qualification burdens. This chapter develops the engineer's working understanding of polymer and metal additive processes, the relationship between process parameters and resulting properties, design for additive (lattices, topology-optimised forms, internal channels), hybrid additive-subtractive workflows, the failure mechanisms that distinguish additive parts (anisotropy, porosity, residual stress, warping), and the qualification approach that bridges single-part novelty to production-grade reliability. The reader who finishes this chapter knows when additive is the right choice and when it is not.

\section{What additive is and is not}\pathtag{core}

Additive manufacturing covers the family of processes that build a part by selectively depositing or fusing material in successive layers. The family is broad. Fused deposition modelling (FDM) extrudes a polymer filament through a heated nozzle along a path planned for each layer. Stereolithography (SLA) cures liquid photopolymer with a laser or projected light. Selective laser sintering (SLS) and multi-jet fusion (MJF) bind powder polymer thermally. Selective laser melting (SLM) and electron-beam melting (EBM) fuse metal powder layer by layer with a focused energy source. Directed energy deposition (DED) blows metal powder or feeds metal wire into a melt pool driven by a laser, electron beam, or arc, building large or repair geometries without a powder bed. Binder jetting prints a liquid binder onto metal or ceramic powder, building a green part that is later sintered.

These processes share a logical workflow. The designer prepares a three-dimensional model. A slicer or build-preparation tool converts the model into layers and toolpaths, with support structures where needed. The machine builds the part. Post-processing follows: support removal, stress relief, hot isostatic pressing for metal, heat treatment, surface finishing, machining of critical features. The post-processing burden distinguishes additive economics from the casual perception of the technology.

What additive can do well: small lot sizes (one to several hundred), complex internal geometry that no subtractive process can reach, lattices and topology-optimised forms, consolidation of an assembly into a single part, rapid iteration of physical prototypes, geometry that follows a function rather than a manufacturing constraint, parts where material cost dominates and the additive process wastes less.

What additive does not do well, despite marketing claims: high-volume production at part costs competitive with injection moulding or die casting; tight tolerances without post-machining; smooth surface finishes without post-finishing; isotropic mechanical properties without careful process control; thin walls in metal without warping; large parts in metal without specialised equipment and long build times; absolute reliability without a qualified process and a qualified part. The novice who designs a part assuming additive is a black box that delivers the CAD geometry is in for surprises.

The decision to use additive is engineering judgement, not enthusiasm. The question is not "can it be printed" but "is additive the right process for this part at this volume with these requirements at this cost." For a single titanium bracket on a satellite where the geometry is topology-optimised and the volume is one, additive is the right answer. For a million plastic widgets at a few cents each, injection moulding is the right answer.

\section{Polymer printing: FDM, SLA, MJF}\pathtag{standard}

FDM (also called fused filament fabrication, FFF) is the workhorse polymer process. A thermoplastic filament is fed through a heated nozzle that traces each layer's perimeter and infill pattern. Layer heights range from $0.05\,\text{mm}$ (slow, fine) to $0.4\,\text{mm}$ (fast, coarse). Common materials include PLA (easy, brittle), ABS (tougher, warps), PETG (chemical resistance), nylon (tough, hygroscopic), polycarbonate (high temperature), and engineering composites filled with carbon or glass fibre.

FDM parts are anisotropic. The bond between layers is weaker than the material within a layer; the print orientation determines which way the part is strong. A bracket loaded in tension perpendicular to layers may have $40$ to $60\%$ of the bulk tensile strength; loaded parallel to layers, $80$ to $95\%$. The designer plans orientation before slicing.

Surface finish on FDM follows the layer height: a $0.2\,\text{mm}$ layer leaves visible steps on curved or angled surfaces. Vertical walls finish best; near-horizontal surfaces (the "stair-step" problem) finish worst. Tolerances of $\pm 0.2\,\text{mm}$ on $100\,\text{mm}$ features are typical; tighter requires post-machining.

SLA cures a liquid photopolymer with light. The build is highly accurate (layer heights of $25$ to $100\,\mu\text{m}$, feature sizes below $0.5\,\text{mm}$) and the surface finish is excellent. The materials are typically brittle, with limited toughness and degraded properties over time and UV exposure. SLA is used for jewellery patterns, dental models, microfluidic prototypes, and visual prototypes where surface finish dominates.

MJF (multi-jet fusion, an HP process) lays down a powder bed, jets a thermal absorber where the part should fuse, and applies infrared energy to fuse the absorbed regions. The result is nearly-isotropic nylon parts with good mechanical properties, surface finish slightly rougher than injection-moulded nylon, and the ability to nest many small parts in a single build. MJF is the polymer process most directly competitive with injection moulding at low to medium volumes ($10$ to $10{,}000$ parts).

\begin{archetype}[scaling: polymer process selection by volume]
At volume $1$ to $10$, FDM or SLA is cheapest. At $10$ to $100$, SLS or MJF often wins. At $100$ to $1000$, MJF or SLS is competitive with low-volume injection moulding (soft tooling). At $1000$ to $10{,}000$, injection moulding with a soft mould usually wins. Above $10{,}000$, hard tool injection moulding is decisive. The volume bands shift with part size, material, and geometric complexity, but the ordering is robust: additive wins low volume, injection moulding wins high volume.
\end{archetype}

\section{Metal printing: SLM, EBM, DED}\pathtag{standard}

Selective laser melting fuses metal powder with a focused laser, building parts in a powder bed under inert gas. Common alloys include Ti-6Al-4V, Inconel 625 and 718, stainless steels (316L, 17-4PH), aluminum alloys (AlSi10Mg), and tool steels. Layer heights are $20$ to $60\,\mu\text{m}$. The resulting density is typically $99.5$ to $99.95\%$ of wrought; with hot isostatic pressing (HIP), $99.99\%$.

Electron-beam melting fuses metal powder with an electron beam under vacuum. The vacuum environment suits reactive alloys (titanium); the higher build temperature reduces residual stress. EBM surface finish is rougher than SLM; resolution is coarser. EBM dominates titanium orthopaedic implants and aerospace components.

Directed energy deposition feeds powder or wire into a melt pool driven by a laser, electron beam, or arc. DED has lower resolution than powder-bed processes but builds large structures (metres) and repairs existing parts. DED is the process behind large-format metal additive in aerospace structures, repair of turbine blades, and additive on existing machined components.

Metal additive parts require post-processing. Stress relief is standard, removing the steep thermal gradients of the build. HIP closes internal porosity (driving density to near-wrought, healing many sub-surface defects). Heat treatment develops the microstructure for the alloy and application. Surface finishing (machining, polishing, abrasive flow, electrochemical) brings critical features to specification. The post-processing burden often exceeds the build time and cost; novices who quote part cost from print-only times underestimate by factors of two to ten.

\begin{estimation}
\textbf{Question.} Estimate the cost per part of an SLM titanium bracket, dimensions $50 \times 50 \times 100\,\text{mm}$, mass $200\,\text{g}$ after machining (build volume includes supports, $\sim 400\,\text{g}$ of titanium powder).

\textbf{Working.} Titanium powder cost: $\sim \$300/\text{kg}$, so material is $\$120$. Machine time: an SLM machine builds $\sim 10\,\text{cm}^3/\text{hr}$; the part plus supports occupy perhaps $\sim 90\,\text{cm}^3$, so build time is $\sim 9$ hours. Hourly rate of an SLM machine (capital, maintenance, gas, labour amortised): $\$50$ to $\$150/\text{hr}$; take $\$100$, giving $\$900$ build cost. Post-processing: heat treat ($\$50$), HIP if required ($\$100$ to $\$300$, share of a batch), wire EDM to separate from build plate ($\$30$), machining of critical features ($\$100$ to $\$300$), surface finishing ($\$50$), inspection (CT scan if required, $\$200$ to $\$500$). Total: roughly $\$1{,}500$ to $\$2{,}500$ per part. The same bracket machined from billet titanium might cost $\$400$ to $\$800$. Additive wins only when topology optimisation cuts the mass, consolidates an assembly, or enables a geometry that machining cannot produce.
\end{estimation}

\section{Process-property relationship}\pathtag{standard}

The mechanical properties of an additive part depend on the process parameters in ways that subtractive manufacturing does not exhibit. The energy density delivered to the powder ($J/\text{mm}^3$) determines whether the powder melts fully, partially, or vaporises. Too little energy: lack of fusion, leaving sub-surface porosity. Too much energy: keyhole porosity from vaporisation, and excessive heat input causing residual stress. The "process window" for a given alloy on a given machine is narrow.

Scan strategy, the order in which the laser traces the layer, controls residual stress. Long continuous scans build up thermal gradients and warping; short scans with alternating directions distribute the heat. Island scanning (subdividing each layer into small squares) is common for large parts.

Powder quality matters in ways that have no analogue in subtractive processes. Powder particle size distribution, sphericity, flowability, and oxygen content all affect the build. Reused powder accumulates oxygen, moisture, and satellite particles; standard practice is to blend reused powder with virgin powder in a tracked ratio. The qualification of a part includes qualification of the powder batch.

The orientation of the part in the build chamber affects properties anisotropically. The build direction is typically weaker than the transverse direction (a function of solidification microstructure and inter-layer bond). HIP and heat treatment reduce but do not eliminate the anisotropy.

Surface finish on metal additive is rough by subtractive standards: as-built $R_a$ of $10$ to $25\,\mu\text{m}$ on flat surfaces, worse on overhangs and down-facing surfaces. Critical features (bearing seats, sealing surfaces, fits) require post-machining. The design must include machining stock on these features.

\section{Design for additive: lattices, internal structures}\pathtag{core}

The capabilities of additive enable design choices that subtractive cannot support. Lattice structures (truss, gyroid, octet) replace solid material with periodic open structures, reducing mass while maintaining stiffness. A $50\%$ relative density gyroid lattice in titanium has roughly $25$ to $30\%$ of the stiffness and $40$ to $50\%$ of the strength of solid titanium, at half the mass. Lattices in implants encourage bone ingrowth; lattices in heat exchangers maximise surface area per volume; lattices in energy-absorbing structures (helmets, crash) tune the force-displacement curve.

Topology optimisation, the computational determination of the optimal material distribution under loads and constraints, produces organic-looking forms that subtractive processes cannot easily make. A topology-optimised satellite bracket may have a mass $30$ to $60\%$ below a conventional design, with internal cavities, curved load paths, and varying cross-sections. The form follows the load path. Designers iterate the optimisation with the manufacturer to ensure the resulting form is printable (overhangs, support requirements, internal channels accessible for powder removal).

Internal channels for cooling, lubrication, or fluid flow are an additive advantage. Conformal cooling channels in injection-moulding tooling, following the contour of the cavity, reduce cycle time and improve part quality. Internal channels in rocket nozzles, turbine blades, and heat exchangers carry coolant where straight-drilled channels cannot reach.

Design rules: maintain overhang angles greater than $45^\circ$ to the build plate (else supports are needed), maintain minimum wall thicknesses ($0.4\,\text{mm}$ FDM, $0.5\,\text{mm}$ SLM, $1\,\text{mm}$ DED), avoid trapped volumes that cannot be drained of unsintered powder, plan support placement and removal access, and include machining stock on critical features.

\begin{principle}[The additive-design covenant]
The geometry that the additive process can build is bounded by physics (overhangs, residual stress, powder removal, support access) and economics (build time, post-processing cost). The designer who treats additive as a geometric free pass discovers these bounds late, usually after a failed build or a part that cannot be cleaned. The competent additive designer learns the process bounds first and designs within them; the topology optimisation runs with the bounds as constraints.
\end{principle}

\section{Hybrid manufacturing}\pathtag{mastery}

Hybrid manufacturing combines additive and subtractive operations, either in a single machine or in a coordinated workflow. A DED head mounted on a CNC milling machine can deposit a feature, machine it to tolerance, deposit another, and so on, building parts that neither additive nor subtractive alone could make. Repair workflows (DED deposit on a damaged turbine blade, followed by machining) are an established hybrid application.

The argument for hybrid: additive provides geometry that subtractive cannot reach (internal channels, near-net shape), while subtractive provides tolerance and surface finish that additive cannot achieve. The argument against: hybrid machines are expensive, complex, and require operators competent in both technologies; the productivity of either operation alone may exceed the hybrid.

For most production parts, the hybrid is realised as a workflow rather than a machine: the part is additively built on one machine, transferred to a second machine for machining, transferred again for surface finishing and inspection. The economics depend on volume, part complexity, and the value of in-process alignment.

\section{Industrial qualification of additive parts}\pathtag{core}

The qualification of an additive part for production differs from the qualification of a machined or cast part in three ways. First, the process is the part: an additive part's properties depend on the build parameters, the powder lot, the machine state, and the post-processing. Each must be qualified, not just the final dimensions. Second, the defect modes are different and harder to detect: sub-surface porosity, lack of fusion, residual-stress-driven cracking, and surface-connected defects that fatigue tests may reveal but visual inspection misses. Third, the historical data set is small: an aerospace component built with twenty years of casting experience is supported by a deep database; the additive version of the same part may have a hundred builds of data.

Qualification standards have emerged. ASTM F2924 (Ti-6Al-4V PBF), ASTM F3001 (Ti-6Al-4V ELI PBF), ASTM F3055 (Ni-718 PBF), ASTM F3091 (polymer PBF), and ISO/ASTM 52900-series cover terminology, qualification methodologies, and material specifications. Aerospace primes have additional internal specifications. FAA AC 33.15-3 and related guidance address aviation use of additive parts.

The qualification approach for a critical additive part: characterise the powder (chemistry, particle size, flowability, oxygen, moisture, batch lot), characterise the machine (energy density calibration, beam offset, build chamber atmosphere, melt pool monitoring), characterise the process (build with test artefacts in known locations, witness specimens for tensile and fatigue testing, CT scan for porosity), characterise the part (CT scan for internal defects, surface inspection, dimensional inspection on critical features, mechanical testing of witness specimens, fatigue testing on production-equivalent geometry), and document the qualification basis. The qualification database grows with each build; the deviations from baseline are tracked and explained.

\section{Failure: anisotropy, porosity, residual stress, warping}\pathtag{core}

The characteristic failure modes of additive parts are different from cast or machined parts.

\textbf{Anisotropy:} mechanical properties differ between the build direction and transverse directions. The inter-layer bond is the weakest direction. Fatigue properties are worst at surface-connected defects, which are more common in the build direction. The designer treats the additive part as an anisotropic material; the FEA model accounts for the build orientation; the orientation is a design parameter, not an afterthought.

\textbf{Porosity:} sub-surface voids from incomplete fusion (lack-of-fusion porosity, large and irregular) or from vaporisation (keyhole porosity, small and spherical). Lack-of-fusion porosity is the more dangerous: it acts as a crack initiator under fatigue loading. CT scanning detects porosity above the resolution limit; HIP closes most porosity but does not heal surface-connected defects. The qualification process tracks porosity statistics by build lot.

\textbf{Residual stress and warping:} the steep thermal gradients during the build leave residual stresses that distort the part during build or after removal from the build plate. Large thin sections are particularly vulnerable. Stress-relief heat treatment is standard. The designer reduces warping by orientation choice, by adding sacrificial structure that constrains the build, and by avoiding large overhangs and long thin features.

\textbf{Surface defects and powder contamination:} the as-built surface roughness, surface-connected porosity, and partially-melted particles on the surface degrade fatigue life. Post-machining, electropolishing, or shot peening can recover fatigue life, but at cost.

\textbf{Powder degradation:} reused powder loses sphericity, gains oxygen, and accumulates contaminants. Builds with degraded powder show porosity, inclusions, and degraded mechanical properties. The powder management protocol (sieving, blending with virgin, oxygen testing, lot tracking) is part of the qualified process.

\textbf{Inconsistent process state:} the machine state (laser power calibration, beam focus, gas flow, oxygen level in the chamber, build plate temperature) drifts. Calibration intervals, in-process monitoring (melt-pool sensors, layer-wise imaging), and witness specimens detect drift. Loss of calibration on a single build can render the parts unfit.

The 2020 Relativity Space and similar aerospace additive programmes operate under formal qualification regimes precisely because the failure modes are real and the cost of escape is high. The novice who treats additive as a print-and-fly technology is a hazard.

\section{Project}\pathtag{core}

\begin{project}[Topology-optimised bracket: design, print, load test]
\textbf{Objective.} Design a small bracket, topology-optimise it for a specified loading condition, print it (FDM polymer, the most accessible option), and load-test it to failure. Compare measured stiffness and strength to FEA prediction; compare mass to a conventional bracket design.

\textbf{Method.} Specify a loading condition (e.g., a $50\,\text{N}$ load at a specified point with a specified attachment geometry). Define the design space as a $50 \times 50 \times 20\,\text{mm}$ envelope with fixed bolt holes. Use a topology-optimisation tool (Altair Inspire, Autodesk Fusion 360 Generative Design, or Ansys Discovery) to produce an optimised geometry. Apply printability constraints. Print the optimised bracket in PLA or PETG with several orientations to test anisotropy. Print a conventional bracket of the same envelope for comparison. Load both to failure on a tensile or three-point bend test, recording force and displacement.

\textbf{Deliverables.} The optimised geometry (CAD file and slicer output), the FEA prediction of stiffness and strength, the printed parts in three orientations, the load test data, a comparison plot of optimised versus conventional, and a written analysis explaining where the prediction matched, where it diverged, and why. Document the anisotropy ratio between strongest and weakest orientation.

\textbf{Hazard.} The load test risks projectile fragments at failure. Use appropriate guarding. Step back from the test stand at the failure load. Wear safety glasses.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
An SLM build of a Ti-6Al-4V part has a melt-pool volume of $0.5\,\text{mm}^3$ per pulse, layer height $30\,\mu\text{m}$, and a build rate of $10\,\text{cm}^3/\text{hr}$. A part of $80\,\text{cm}^3$ (including supports) is built. Compute the build time.
\end{exercise}

\begin{exercise}
An FDM bracket is printed with $0.2\,\text{mm}$ layer height, $40\,\text{mm/s}$ print speed, $0.4\,\text{mm}$ nozzle, and $20\%$ infill. The bracket envelope is $100 \times 50 \times 20\,\text{mm}$. Estimate the print time and the material mass (PLA density $1.24\,\text{g/cm}^3$).
\end{exercise}

\begin{exercise}
A topology-optimised SLM titanium bracket has mass $150\,\text{g}$, replacing a $400\,\text{g}$ machined-from-billet bracket. The mission is a satellite at $\$10{,}000/\text{kg}$ launch cost. Compute the launch-cost saving per bracket. State whether the additive cost premium of $\$1500$ per bracket is justified.
\end{exercise}

\begin{exercise}
An SLM Ti-6Al-4V part has $0.5\%$ porosity by volume, average pore size $50\,\mu\text{m}$. Estimate the reduction in fatigue strength compared to wrought (use a simple Murakami-type estimate or state the model and parameters).
\end{exercise}

\begin{exercise}
A laser scan path has $200\,\text{W}$ power, $1000\,\text{mm/s}$ speed, $30\,\mu\text{m}$ layer height, $80\,\mu\text{m}$ hatch spacing. Compute the energy density delivered to the powder ($J/\text{mm}^3$).
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the relationship between scan-strategy island size and the peak residual stress in a flat plate build, using a simple thermal-gradient model. State the assumptions and the regime where the model applies.
\end{exercise}

\begin{exercise}
Derive the stiffness of a gyroid lattice as a function of relative density, using a beam-bending model for the unit cell. State the bounds on the model.
\end{exercise}

\begin{exercise}
Show that an SLM build's surface roughness scales with layer height for vertical walls but with $1/\sin\theta$ for down-facing surfaces at angle $\theta$. Sketch the geometry.
\end{exercise}

\begin{exercise}
Derive a simple relationship between melt-pool depth and laser power, speed, and absorptivity, using a steady-state heat-conduction model. Estimate the melt-pool depth for $200\,\text{W}$, $1000\,\text{mm/s}$ on titanium.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the cost per part of a hundred-piece batch of FDM PLA brackets, envelope $80 \times 40 \times 10\,\text{mm}$, $0.2\,\text{mm}$ layers, $20\%$ infill. State assumptions about machine cost, throughput, and labour.
\end{exercise}

\begin{exercise}
Estimate the mass reduction achievable by topology optimisation versus a conventional milled bracket for a satellite. State assumptions about loading, factor of safety, and material choice.
\end{exercise}

\begin{exercise}
Estimate the porosity that can be tolerated in an SLM aerospace bracket fatigue-loaded to $200\,\text{MPa}$, with a fatigue limit at $400\,\text{MPa}$ for defect-free material. State the model.
\end{exercise}

\begin{exercise}
Estimate the heat treatment time and energy for stress-relief of an SLM titanium part of $5\,\text{kg}$ mass.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Model the residual stress distribution in a flat SLM plate using a thermal-elastic FEA. Compare three scan strategies (continuous, island, alternating) and report peak stress for each.
\end{exercise}

\begin{exercise}
Simulate the cooling-rate distribution in an SLM build for a thin wall and a thick wall, on the same plate. Identify the regions most likely to crack.
\end{exercise}

\begin{exercise}
Model the elastic stiffness of a gyroid lattice at three relative densities ($30\%$, $50\%$, $70\%$). Compare to a closed-form estimate.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a topology-optimised bracket for a $200\,\text{N}$ load at a single point, fixed at three bolt holes, envelope $80 \times 80 \times 20\,\text{mm}$, target mass reduction $40\%$ from a conventional design.
\end{exercise}

\begin{exercise}
Design a manifold with three internal channels, no external machining, suitable for SLM titanium. State the wall thicknesses, the support strategy, and the powder-removal path.
\end{exercise}

\begin{exercise}
Design a lattice-filled heat exchanger that maximises surface area per volume while maintaining $1\,\text{mm}$ minimum wall and allowing the working fluid to flow.
\end{exercise}

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A team proposes to additively manufacture a flight-critical bracket in titanium. The team has built and tested twelve parts. State the engineering questions to ask before approving production.
\end{exercise}

\begin{exercise}
A vendor claims that SLM parts are equivalent to wrought after HIP. Critique the claim. State the residual differences.
\end{exercise}

\begin{exercise}
A team finds that two builds of the same part show $\pm 15\%$ variation in tensile strength. State three engineering hypotheses and the corresponding investigations.
\end{exercise}

\begin{exercise}
A team's additive part fails fatigue testing at $30\%$ of the wrought fatigue limit. State three likely causes and the corresponding mitigations.
\end{exercise}

\begin{exercise}
A team is considering replacing a machined aluminum bracket (production volume $10{,}000/\text{year}$) with an additive titanium bracket. State the engineering trade-offs and conditions under which the replacement is justified.
\end{exercise}

\begin{exercise}
A team uses recycled powder for cost reasons; the most recent build shows higher porosity than baseline. State the engineering investigation, the diagnostic measurements, and the corrective actions to consider before the next build.
\end{exercise}
