\chapter{Ethics under risk and uncertainty}
\label{vol11:ch11}

\epigraph{The engineering profession's ethics are not the supplement to its technical work; they are the discipline within which the technical work is done.}{}

\chapmeta{Half-life: durable. Archetypes: ethics. Exercise target: 15. Project track: hybrid.}

\noindent
Engineering ethics is treated in this chapter as a constitutive feature of design rather than as a postscript. The engineer who exercises judgment under risk and uncertainty is making ethical commitments in every design decision: what risk to accept; what evidence to demand; what populations to serve; what populations to exclude; what to disclose; what to suppress; what to refuse. The chapter develops the discipline: how the codes of ethics translate into working practice; what the duties to report and to refuse require; how conflicts of interest are managed; where the line lies on dual-use, surveillance, and military work; what whistleblowing demands and what protects it; how ethical compromises become technical disasters.

\section{Engineering ethics as practice, not aspiration}\pathtag{core}

Engineering ethics has often been taught as aspiration: the engineer should care about the public welfare; the engineer should be honest; the engineer should be responsible. The aspiration is correct as far as it goes; it under-specifies the discipline that turns the aspiration into action. The practice-oriented framing of this chapter treats ethics as decisions made in specific design situations, with specific consequences, and with specific tools the engineer can use.

The substantive ethical questions in engineering work include:

\textbf{How much risk to accept.} Designs have residual risk; the risk is borne by users, by third parties, by the public, by future populations. The engineer's risk-acceptance decisions are ethical decisions: who is exposed to what risk for what benefit, and whether the exposure is informed.

\textbf{What evidence to demand.} Designs are committed under uncertainty; the engineer decides how much evidence is enough. The decision is ethical: insufficient evidence exposes others to consequences the engineer did not adequately characterise.

\textbf{Which populations to serve.} Designs serve some populations and not others. The engineer's decisions about whose needs to address are ethical decisions: the populations that are not served bear the cost of their exclusion.

\textbf{What to disclose.} Designs have limitations, failure modes, residual risks. The engineer's decisions about what to disclose to users, to regulators, to the public are ethical decisions: the parties not informed cannot make appropriate decisions.

\textbf{What to refuse.} Some designs the engineer is asked to participate in are designs the engineer should refuse to be part of. The refusal is the ethical exercise of professional judgment.

\textbf{When to escalate.} Some concerns the engineer encounters during work warrant escalation beyond the immediate chain of command. The escalation decision is ethical: the cost of escalation to the engineer's career is weighed against the cost to others of the un-escalated concern.

The ethical discipline includes the discipline of recognising these decisions as ethical: many engineers make them daily without recognising the framing; the recognition is the precondition for the discipline.

The codes of ethics of the engineering professional bodies (NSPE Code of Ethics; IEEE Code of Ethics; ASCE Code of Ethics; ASME Code of Ethics; the various international equivalents) provide the canonical framing. The codes share a common structure: the engineer's first duty is to the public's safety, health, and welfare; subsequent duties are to the client, the profession, and the engineer's own development. The codes are read by engineers in training, are tested in professional registration examinations, and are enforced (with varying rigour) through professional discipline processes.

The codes have limits. They state principles but do not always specify their application to particular cases; they require enforcement structures that vary in effectiveness across jurisdictions; they apply most strongly to licensed professional engineers and less strongly to engineers without professional licensure (which is most engineers in many sectors). The codes are the substantive starting point of engineering ethics; the discipline of applying them is the substantive engineering work.

\begin{principle}[Ethics is in every decision]
Engineering ethics is not a separate domain from engineering work; it is the framing within which the work is done. Every design decision involves judgments about risk, evidence, populations served, disclosure, and refusal; every judgment is an ethical commitment whose consequences fall on parties beyond the engineer. The discipline of engineering ethics is the discipline of making the judgments visible, the commitments explicit, and the consequences accountable.
\end{principle}

\section{Codes of ethics and what they actually do}\pathtag{standard}

The major engineering codes of ethics share a common structure:

\textbf{Paramount duty to the public.} The first canon of the NSPE code states "Engineers shall hold paramount the safety, health, and welfare of the public." The IEEE code's first principle states "to hold paramount the safety, health, and welfare of the public." The ASME code's first canon states "Engineers shall hold paramount the safety, health, and welfare of the public in the performance of their professional duties." The wording is similar across codes; the priority is consistent.

\textbf{Honesty in claims and representations.} The codes require honesty in statements made in professional capacity: only signing what the engineer competently affirms; not making knowingly false claims; not misleading on the basis or limits of professional work.

\textbf{Acting within competence.} The codes require engineers to undertake only work within their competence; the engineer who works outside competence is expected to acquire additional competence or to engage others with the competence.

\textbf{Loyalty to client / employer with limits.} The codes require professional loyalty to the client or employer, but the loyalty has explicit limits: the loyalty does not override the duty to the public's safety; the loyalty does not require dishonesty; the loyalty does not require participation in unsafe or unethical work.

\textbf{Disclosure of conflicts of interest.} The codes require disclosure of conflicts of interest to the affected parties.

\textbf{Confidentiality.} The codes require maintenance of confidentiality regarding the client's or employer's proprietary information, with explicit exceptions for safety and legal matters.

\textbf{Professional development.} The codes require continuing professional development; the engineer's competence is maintained through ongoing learning.

\textbf{Acting in professional manner.} The codes require professional conduct: civility in disagreement; constructive engagement with colleagues; respect for others' contributions; appropriate credit.

The codes do not address every ethical situation; they are general principles to be applied with judgment. The professional bodies maintain Board of Ethical Review or equivalent structures that publish opinions on specific cases; the opinions are the discipline's case law and provide guidance on application.

The codes' enforcement varies:

\textbf{Licensed engineers.} The professional licensure body can investigate complaints, conduct hearings, and impose sanctions up to revocation of the licence. The sanctions are real; loss of licence is career-affecting.

\textbf{Society members.} The professional society can investigate complaints and impose sanctions including expulsion from the society. The sanctions are less consequential than licensure sanctions but produce professional reputation effects.

\textbf{Non-licensed, non-society engineers.} The codes apply as professional norms but enforcement depends on employer policies, criminal liability, civil liability, and reputational consequences rather than on direct professional sanctions.

The codes' actual influence on day-to-day engineering decisions varies. In domains with strong licensure traditions (civil engineering; mechanical engineering in some industries; certain medical-device contexts), the codes are referred to and considered explicitly. In domains with weak licensure traditions (software engineering in most contexts; consumer-product engineering; many startup contexts), the codes are less salient and the ethical decisions are made implicitly. The discipline of explicit ethical engagement is partly a matter of professional culture; the cultures vary by domain and by organisation.

\section{The duty to report and the duty to refuse}\pathtag{core}

Two specific duties are central to the practice-oriented framing of ethics: the duty to report and the duty to refuse.

\textbf{The duty to report.} The engineer who learns of facts that bear on the public's safety, health, or welfare has a duty to report the facts to parties who can act on them. The duty includes:

Reporting safety issues in designs the engineer encounters in professional work (the design has a hazard the design organisation has not addressed; the design's safety analysis is inadequate; the design's operating envelope does not match the service conditions). The first report is typically internal: to the design organisation's leadership, to the safety function, to the project management. If the internal report does not result in adequate action, the duty extends to external reporting: to the regulator (where one exists); to the relevant professional body; in extreme cases, to public disclosure.

Reporting violations of professional standards by other engineers when the violations bear on public safety. The duty is to the public's protection, not to the colleague's discomfort; the discomfort is a real cost but does not override the duty when the public's safety is at stake.

Reporting conditions that would affect the engineer's signed design's continued safety: changes in operating environment; deterioration the engineer observes; new failure modes the engineer learns of. The reports support the design's continued safe operation through its service life.

\textbf{The duty to refuse.} The engineer asked to participate in work the engineer believes is unsafe or unethical has a duty to refuse. The refusal includes:

Refusing to sign designs the engineer does not believe are adequate. The signature commits the engineer to the design's adequacy (Chapter 9); the engineer who signs designs the engineer does not believe are adequate has violated the signature's substance.

Refusing to perform analyses the engineer knows will be misused. An analysis conducted to justify a pre-decided conclusion is not an analysis; the engineer who knowingly performs such analyses has participated in the misuse.

Refusing to participate in work the engineer believes is unethical regardless of the work's safety: weapons systems the engineer believes should not exist; surveillance systems the engineer believes will be misused; environmentally destructive projects the engineer believes are not justified. The refusal is a professional judgment; the engineer bears the consequences of the refusal (career impact; commercial impact) and the consequences are the cost of the discipline.

Refusing to violate the professional codes regardless of organisational pressure to do so. The engineer who is asked to falsify data, to suppress findings, to deceive regulators or customers has a duty to refuse; the refusal may produce career consequences but the alternative is professional violation.

The duties to report and refuse are exercised against organisational and economic pressure that typically opposes them. The codes of ethics protect the engineer in form (the engineer who acts in accordance with the codes cannot be professionally disciplined for the action); the protection in practice depends on the engineer's licensure, the organisation's ethical culture, the regulatory framework, the legal protections for whistleblowers. The exercise of the duties is often professionally and personally costly; the alternative is the violation the duties are designed to prevent.

\begin{archetype}[ethics]
The duties to report and to refuse are the ethics archetype's working expression. The engineer's professional judgment about specific designs and specific work produces ethical commitments the engineer must act on; the action is reporting when the issue can be addressed by others, and refusing when participation would itself be the issue. The discipline is exercised under pressure; the discipline's institutional protections vary; the engineer's commitment to the discipline is what distinguishes the profession from the work.
\end{archetype}

\section{Conflicts of interest}\pathtag{standard}

A conflict of interest exists when the engineer has personal, financial, or professional interests that could affect the engineer's professional judgment. The conflict's existence does not itself make the engineer's judgment wrong; the engineer's response to the conflict determines whether the judgment is sound.

The categories of conflict:

\textbf{Financial interests.} The engineer owns equity in a company that benefits from the design decision; the engineer holds a consulting contract with a party affected by the decision; the engineer's compensation depends on the decision's outcome.

\textbf{Personal relationships.} The engineer has a family member, friend, or close colleague who is affected by the decision. The relationships may affect the engineer's judgment even if the engineer is not consciously aware.

\textbf{Professional relationships.} The engineer's prior or ongoing professional work creates relationships with parties affected by the decision: prior employers; clients; collaborators; competitors.

\textbf{Reputational interests.} The engineer's professional reputation is enhanced or threatened by the decision's outcome: a successful design enhances reputation; a failed design damages it; the prospect of either can affect judgment.

\textbf{Future-employment prospects.} The engineer's prospective employer (or client) is affected by the engineer's current decision; the prospect of future engagement may affect the engineer's judgment.

The response to conflicts:

\textbf{Disclosure.} The conflict is disclosed to the parties affected by the engineer's decision: the employer; the client; the regulator; in some cases, the public. Disclosure allows the affected parties to evaluate the engineer's judgment in light of the conflict.

\textbf{Recusal.} The engineer steps back from decisions the conflict could affect; the decision is made by parties without the conflict. The discipline is the substantive response to substantive conflicts.

\textbf{Mitigation.} The conflict's effect on the decision is mitigated through structural means: independent review of the engineer's work; explicit decision criteria that the conflict cannot subjectively influence; involvement of multiple parties whose conflicts cancel.

\textbf{Avoidance.} Some conflicts are avoided in advance: the engineer does not accept work that creates the conflict; the engineer divests financial interests that would create the conflict; the engineer maintains separation between personal relationships and professional decisions.

The discipline of conflict-of-interest management is the discipline of recognising the conflict (often the recognition is the harder step), of responding appropriately, and of maintaining the response over time as circumstances change. The codes of ethics require disclosure; the codes' enforcement against undisclosed conflicts is one of the more reliably enforced aspects of professional discipline.

\section{Dual-use, surveillance, military: where the line is}\pathtag{standard}

Some engineering work is dual-use: it can be used for purposes the engineer agrees with and for purposes the engineer does not. Some engineering work is unambiguously military or surveillance-related. The engineer's decisions about participating in such work are ethical decisions; the codes of ethics give limited specific guidance and the substantive judgment is the engineer's.

The categories:

\textbf{Dual-use civilian technology.} Most engineering technology is dual-use to some degree: the chemistry that produces fertilisers also produces explosives; the cryptography that protects privacy also protects criminal communications; the navigation that guides commercial aircraft also guides military aircraft. The engineer's role in dual-use technology is partly the design's intended use (the engineer's primary commitment) and partly the design's foreseeable misuse.

\textbf{Surveillance technology.} Designs explicitly intended to monitor populations: facial recognition; metadata analysis; tracking systems; predictive policing. The technology can serve legitimate purposes (criminal investigation; public safety) and can be misused (mass surveillance of populations; suppression of dissent; targeting of minorities). The engineer's judgment is about the technology's likely deployment context and the engineer's willingness to contribute to the likely outcomes.

\textbf{Military weapons systems.} Designs explicitly for military use: weapons; targeting systems; military communications; command-and-control. The engineer's judgment is about the legitimacy of the military activity and the engineer's willingness to participate in the violence the weapons enable.

\textbf{Privacy-invasive consumer technology.} Designs that collect more personal data than necessary, that share data without informed consent, that create surveillance infrastructure as a side effect of consumer convenience. The engineer's judgment is about whether the design's data practices serve users or extract value from them.

\textbf{AI systems with consequential outcomes.} Designs that make decisions affecting individuals: credit scoring; hiring; medical diagnosis; criminal justice. The engineer's judgment is about the system's accuracy, fairness, transparency, and accountability.

The engineering professional bodies have grappled with these categories without producing settled positions. The IEEE has published guidance on AI ethics; the major engineering societies have engaged on military and surveillance technology; the discussion is ongoing and the codes evolve. The substantive judgment is the engineer's; the codes provide framing rather than determination.

The lines that individual engineers draw vary:

Some engineers participate in any legal engineering work; their professional commitment is to engineering quality regardless of the use.

Some engineers refuse military work; their professional commitment includes a refusal to contribute to violence.

Some engineers refuse surveillance work; their professional commitment includes a refusal to contribute to population monitoring.

Some engineers refuse work on systems they believe are biased or discriminatory; their professional commitment includes a refusal to contribute to harm.

Some engineers engage in dual-use technology with explicit constraints on their work: contributing to civilian applications but not military ones; contributing to development but not deployment; contributing to research but not production.

The engineering profession does not have a single answer; the answers individual engineers reach are part of their professional identity. The discipline is to make the answers explicit, to articulate the reasoning, and to act in accordance with the articulated position rather than drifting into participation that the engineer would on reflection refuse.

\section{Whistleblowing}\pathtag{standard}

Whistleblowing is the disclosure of internal misconduct to external parties: the press; the regulator; a legal authority; a public-interest organisation. The engineer who whistleblows has typically exhausted internal channels for raising the concern; the external disclosure is the last-resort exercise of the duty to report.

The conditions that produce whistleblowing:

\textbf{Internal channels have failed.} The engineer raised the concern internally; the response was inadequate or absent; the issue continues unaddressed.

\textbf{The issue is substantive.} The misconduct is not a personal grievance or a disagreement about policy; it is conduct that bears on public safety, on legal compliance, on financial integrity, on professional standards.

\textbf{The disclosure is necessary.} The external disclosure is necessary to address the issue; less drastic interventions have been tried or are clearly inadequate.

\textbf{The engineer accepts the personal cost.} Whistleblowing typically produces severe personal cost: career damage, professional retaliation, legal exposure, social ostracism. The engineer makes the disclosure knowing the cost.

The protections for whistleblowers:

\textbf{Legal protections.} Many jurisdictions have whistleblower-protection laws: the US Whistleblower Protection Act; the UK Public Interest Disclosure Act; the EU Whistleblower Directive; the various national equivalents. The laws prohibit retaliation against whistleblowers and provide legal remedies; the protection's substance varies by jurisdiction and by the type of disclosure.

\textbf{Sector-specific protections.} Some sectors have additional protections: aviation safety reporting through ASRS (Volume X Chapter 13); financial-services whistleblower programmes; medical-error reporting systems; nuclear-industry safety reporting. The sector-specific structures often provide stronger protections than the general laws.

\textbf{Professional-body support.} The engineering professional bodies provide support to whistleblowers in some cases: ethics consultations; expert testimony; financial assistance with legal proceedings.

\textbf{Public-interest organisations.} Organisations focused on whistleblower support (Government Accountability Project; National Whistleblower Center; various sector-specific organisations) provide guidance, legal assistance, and public attention.

The institutional realities:

Whistleblowers face severe consequences even with legal protections. Retaliation is sometimes overt (firing; demotion; reassignment to undesirable work) and sometimes covert (exclusion from projects; negative performance reviews; pressure from colleagues). The legal protections require enforcement; enforcement varies in effectiveness.

The whistleblower's claims often face contestation: the employer disputes the facts; the regulator's response is slow; the public discussion mixes the substantive issue with the personality conflict. The whistleblower's burden includes the substantive work of articulating the issue and the personal work of withstanding the contest.

The whistleblower's choice is between participating in the misconduct (or remaining silent about it) and accepting the personal cost of disclosure. The choice is ethical; the codes of ethics support the disclosure but cannot eliminate the cost.

The Volume X cases include whistleblower-relevant patterns: the Roger Boisjoly testimony after Challenger; the FAA authorised representatives' concerns about 737 MAX MCAS that were inadequately escalated; the Volkswagen engineers who knew of the defeat device but did not disclose it. The pattern is recurring: organisations with active whistleblower channels have better safety records; organisations that suppress whistleblowing have worse ones.

\section{Failure: ethical compromises that became technical disasters}\pathtag{core}

The Volume X cases are largely failures whose technical mechanisms are well-documented and whose ethical mechanisms are equally substantive. The pattern: ethical compromises at the design or management level produced the conditions under which the technical failures occurred.

\textbf{Challenger.} The launch decision was made over the explicit objections of Morton Thiokol engineers who had documented the O-ring's vulnerability to low temperature \cite{acc:nasa-challenger}. The Thiokol management's reversal of their engineers' recommendation was an ethical compromise (the management chose to accept the risk against the engineers' judgment); the compromise produced the conditions under which the technical failure occurred. The post-accident reforms addressed the technical mechanism (O-ring redesign; field-joint redesign); the ethical conditions (the relationship between engineering and management; the channels for engineering escalation; the consequences of unheeded escalation) were addressed through the institutional reforms but the underlying pattern has recurred.

\textbf{Columbia.} The engineers who recognised that the foam strike might have caused critical damage requested on-orbit imagery; the program management did not authorise the request \cite{acc:caib-columbia}. The denial was an ethical compromise (the management chose to accept the risk against the engineers' judgment); the compromise produced the conditions under which the technical failure could not be diagnosed before re-entry. The CAIB report's findings on the parallels with Challenger's ethical mechanism are explicit.

\textbf{737 MAX MCAS.} The certification process was conducted under the FAA's ODA arrangement; the Boeing engineers and authorised representatives raised concerns about MCAS that were inadequately escalated \cite{acc:house-737max-2020}. The House Committee report identifies specific ethical compromises: the commercial premise that the MAX would require no simulator training shaped what could be documented to operators; the AoA-disagree-alert non-installation was not disclosed to the FAA for over a year; the safety-assessment was not rerun when MCAS's authority was expanded. The technical failure mode (single-AoA architecture for a function with substantial authority) was the consequence of the ethical compromises.

\textbf{Volkswagen diesel emissions.} The defeat device was implemented by engineers who knew it was regulatory evasion \cite{acc:epa-vw-nov-2015}. The ethical compromise (the engineers participated in the implementation; the management directed it; no internal whistleblower disclosed it) produced the regulatory and commercial consequences that followed. The compromise's scale (the engineering hours that went into implementing and maintaining the defeat device) reflects the substantive participation across multiple engineers.

\textbf{Bhopal.} The pre-accident plant's safety systems had been progressively disabled to save operating cost; the safety conditions that the design specified were not maintained in operation \cite{acc:icmr-bhopal-2004}. The ethical compromise (the operator's choice to operate the plant without the specified safety provisions; the parent corporation's tolerance of the deteriorated state) produced the conditions under which the technical accident occurred.

The pattern recurs across the Volume X cases: the technical mechanism is the proximate cause; the ethical mechanism is the structural cause. The engineering response that addresses only the technical mechanism (the O-ring redesign; the MCAS architecture revision; the emissions-testing protocols) is incomplete; the engineering response that also addresses the ethical mechanism (the engineering-management relationship; the safety-engineering authority; the whistleblower channels) is more substantive.

The chapter's discipline is the discipline of recognising the ethical mechanism in current design work, before it becomes the technical disaster of the next case. The recognition is partly the engineer's individual discipline (this chapter's framing) and partly the institutional support (the codes of ethics; the licensing structures; the whistleblower protections; the regulatory frameworks). The combination is what the engineering profession can offer; the alternative is the recurrence of the Volume X patterns.

\begin{principle}[Ethical compromises produce technical disasters]
The Volume X cases consistently show that ethical compromises at the design and management level produce the conditions under which technical failures occur. The engineering response to past failures has often addressed the technical mechanism without addressing the ethical mechanism; the patterns have recurred. The discipline of design includes the discipline of recognising and resisting ethical compromises before they produce the technical disasters that the post-failure analysis will subsequently identify.
\end{principle}

\begin{project}[Personal ethics statement]
Write a personal engineering-ethics statement: a 2-3 page document that defines (a) the work the reader will do; (b) the work the reader will not do; (c) the reasoning behind the boundaries; (d) the conditions under which the reader would re-examine the boundaries; (e) the institutional supports the reader would rely on in difficult ethical situations. The statement is personal but disciplined: it is grounded in the codes of ethics this chapter has developed; it is specific to the reader's domain and circumstances; it is defensible against substantive challenge.

\textbf{Hazard.} The statement is a personal commitment; writing it is itself a substantive ethical exercise. The point is the discipline of explicit articulation rather than the production of a polished document. The reader is encouraged to revisit and revise the statement periodically over their career as circumstances and judgment evolve.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
The NSPE Board of Ethical Review has published approximately 500 case opinions over its existence. Estimate the average rate per year. Discuss the implications for the discipline's living-case-law character and the engineer's access to applied ethical guidance.
\end{exercise}

\begin{exercise}
A whistleblower who is fired in retaliation for their disclosure may pursue legal remedies; the average litigation takes approximately 3 years and costs \$200{,}000 to \$500{,}000. Compute the financial exposure relative to the median engineering salary. Discuss the practical implications for whistleblower decisions.
\end{exercise}

\begin{exercise}
The engineering professional bodies report annual disciplinary actions in the tens to hundreds across the major licensing jurisdictions; the actions affect a small fraction of the licensed engineer population. Compute the implied annual rate of disciplinary action per licensed engineer. Discuss what the rate implies about the enforcement of the codes.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive a simple model of the engineer's incentive structure for raising versus suppressing safety concerns. Assume: raising a concern has cost $C_R$ (time; career; relationship); suppressing has cost $C_S$ (probability $p$ of consequence; ethical cost $E$). Identify the conditions under which each is preferred and the institutional levers that shift the balance.
\end{exercise}

\begin{exercise}
For an organisation with $N$ engineers, each independently encountering ethical concerns at rate $\lambda$ and reporting with probability $r$, derive the expected number of reports per year. Identify the conditions under which the reporting rate is sufficient to surface systemic patterns versus the conditions under which it is too low to be informative.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the fraction of major engineering failures (from the Volume X cases) where the post-accident investigation identified ethical compromises in the causal chain. Discuss the implication for engineering education and professional practice.
\end{exercise}

\begin{exercise}
Estimate the personal cost to a whistleblower (salary loss; legal costs; reputational impact; emotional cost) over the 5 years following a disclosure that leads to retaliation. Compare to the public benefit if the disclosure prevents a comparable failure to the Volume X cases.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design an internal ethics-concern-reporting system for an engineering organisation. Specify: the channels; the protections; the escalation paths; the response standards; the public-disclosure protocols for substantive findings.
\end{exercise}

\begin{exercise}
Design an ethics-review protocol for design reviews. Include specific questions about: who bears the design's residual risk; what disclosure obligations the design creates; what populations are served and not served; what dual-use considerations apply.
\end{exercise}

\begin{exercise}
Design an organisational policy on dual-use technology (military, surveillance, AI-with-consequential-outcomes) for a hypothetical engineering firm. Specify the work the firm will and will not undertake, the criteria for individual engineer participation, and the resolution of disagreement.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
A junior engineer believes a senior engineer has approved a design that contains a safety issue. Discuss the junior engineer's options under the codes of ethics. Identify the institutional structures that support or undermine the junior engineer's substantive engagement with the issue.
\end{exercise}

\begin{exercise}
Argue: the engineering profession's codes of ethics provide insufficient practical guidance on dual-use technology; the codes need to be strengthened with sector-specific guidance for AI, surveillance, and military work. Argue the opposite: general codes plus individual engineer judgment is sufficient. Identify the empirical evidence that bears on the dispute.
\end{exercise}

\begin{exercise}
Discuss the Volkswagen diesel-emissions case as an ethical-compromise pattern. Identify the institutional and individual mechanisms by which the engineers' participation was sustained over years. Discuss what would have been required to disrupt the pattern.
\end{exercise}

\begin{exercise}
An engineer is offered substantially higher compensation to work on a project the engineer believes raises ethical concerns. Discuss the engineer's options for evaluating the offer: the codes of ethics framing; the personal values framing; the long-term professional reputation framing. Identify the structural conditions that make the decision difficult.
\end{exercise}

\begin{exercise}
The engineering profession in many domains has weak licensure (software; consumer products; many sectors) and consequently weak institutional enforcement of ethical codes. Discuss the implications for engineering ethics as actually practised. Identify the alternative structures (employer policies; legal liability; reputation; collective action) that compensate or fail to compensate.
\end{exercise}
