\chapter{Vectors and vector calculus}
\label{vol02:ch08}

\epigraph{Vector analysis is the language in which the local and the
global are joined: a quantity that flows through a point and a quantity
that accumulates inside a region are the same statement, written from
two ends of an integral.}{The working engineer's gloss on the
divergence theorem, after \textcite{text:strang2016}.}

\chapmeta{Half-life: foundational. Archetypes: balance (the divergence
theorem as the formal accounting equation between an interior source
and a surface flux); transport (the gradient as the local driver of
flux, the divergence as the local rate of net outflow per unit volume).
Exercise target: 35. Project track: household + analysis (hybrid).
Hazard class: none. Reader path default: standard, with sections
explicitly tagged. Named cases: none required; the failure section
treats orientation and sign-of-the-normal errors generically.}

\noindent
A vector is a quantity that carries both magnitude and direction;
vector calculus is the differential and integral apparatus by which
vector-valued functions of position are described, manipulated, and
balanced. The reader has met directed quantities since Volume I:
displacement, velocity, force, electric current, the gradient of a
temperature field. The mathematics of this chapter formalises a
practice the reader already has, and turns three common physical
operations, gradient, divergence, and curl, into a single bookkeeping
discipline. The chapter closes with the three integral theorems of
Green, Gauss, and Stokes, which bind the local rate of change of a
field to the integrated boundary behaviour of that field, and with
the project that earned the chapter its place in the volume: the
divergence theorem applied as a volume calculation against a sampled
surface, reconciled with the empirical water-displacement result the
reader produced in Volume~I, Chapter~7.

\section{Vectors as algebraic and geometric objects}\pathtag{core}

A \engterm{vector} in $\R^{n}$ is an ordered tuple
$\vect{v} = (v_{1}, v_{2}, \ldots, v_{n})$ together with the
operations of componentwise addition and scalar multiplication. We
write the same object as a column,
\[
\vect{v} = \begin{pmatrix} v_{1} \\ v_{2} \\ \vdots \\ v_{n} \end{pmatrix},
\]
and use the boldface convention $\vect{v}$ throughout this volume; the
older $\vec{v}$ notation appears in handwritten work, in some primary
sources, and in figure annotations. Both denote the same object. A
scalar, by contrast, is set in upright italic, $v$, and represents a
single real number.

The geometric content is older than the notation. A vector represents
a directed segment with a tail and a head; the segment's length is the
\engterm{magnitude} or \engterm{norm} $\norm{\vect{v}}$, and its
direction is the unit vector $\unitvec{v} = \vect{v} / \norm{\vect{v}}$.
The two pictures, algebraic tuple and geometric arrow, are connected
by a choice of \engterm{basis}: a set of $n$ linearly independent
vectors $\vect{e}_{1}, \ldots, \vect{e}_{n}$ such that any $\vect{v}$
is uniquely written as
$\vect{v} = v_{1}\vect{e}_{1} + \cdots + v_{n}\vect{e}_{n}$.
The Cartesian basis $\unitvec{x}, \unitvec{y}, \unitvec{z}$ in $\R^{3}$
is the working default, and the components $v_{i}$ in that basis are
the numbers an engineer reads off an instrument or writes on a drawing.

\subsection{Operations and identities}

Vector addition is componentwise, $(\vect{u} + \vect{v})_{i} = u_{i} +
v_{i}$, and obeys the parallelogram law geometrically: place
$\vect{u}$ and $\vect{v}$ tail to tail; their sum is the diagonal of
the parallelogram they span. Scalar multiplication
$\alpha \vect{v} = (\alpha v_{1}, \ldots, \alpha v_{n})$ stretches
the arrow by $\abs{\alpha}$ and reverses it when $\alpha < 0$.
Subtraction is addition of the negation: $\vect{u} - \vect{v} =
\vect{u} + (-1)\vect{v}$ points from the head of $\vect{v}$ to the
head of $\vect{u}$ when both share a tail.

The norm $\norm{\vect{v}} = \sqrt{v_{1}^{2} + \cdots + v_{n}^{2}}$
satisfies the three axioms a length must satisfy: it is non-negative,
it is zero only for the zero vector, and it obeys the triangle
inequality $\norm{\vect{u} + \vect{v}} \leq \norm{\vect{u}} +
\norm{\vect{v}}$ \cite{text:strang2016}. The triangle inequality is
the algebraic statement that the shortest path from a tail to a head
is a straight line; it becomes the engineer's tool for bounding
errors in vector-valued sums.

\subsection{Why vectors, not scalars}

A force on a beam, a velocity in a pipe, and an electric field at a
point all carry direction. A scalar description loses information the
engineer needs. A force of $100 \,\si{\newton}$ is not the same load
on a column as $100 \,\si{\newton}$ tangent to the column's axis; the
first is a compression, the second is a shear. A velocity of
$10 \,\si{\meter\per\second}$ in a duct is not the same air supply as
$10 \,\si{\meter\per\second}$ at $80^{\circ}$ to the duct axis; the
first delivers airflow, the second delivers swirl. The engineering
record is built from vectors precisely because the algebra of scalars
hides the directional information that determines whether a system
carries its load, delivers its mass flow, or aligns its sensor.

\begin{archetype}[Transport, vector form]
A transport quantity is one that flows from one region to another:
mass, momentum, charge, energy, signal. Its local description is a
\engterm{flux density}, a vector $\vect{J}$ whose magnitude is the
amount transported per unit area per unit time and whose direction is
the direction of net transport. The product $\vect{J} \cdot \unitvec{n}
\,dA$ is the rate of transport across an oriented surface element.
The chapter's later sections build this transport accounting into the
divergence theorem, but the vector $\vect{J}$ is the irreducible
object: a directed rate per unit area at a point.
\end{archetype}

\begin{estimation}
\textbf{Question.} A garden hose delivers water to a household lawn at
a measured volumetric flow rate of $8 \,\si{\liter\per\minute}$
through a nozzle of inner diameter $12 \,\si{\milli\meter}$. Estimate
the water's exit velocity vector at the nozzle, and the magnitude of
the velocity at a point on the lawn where the jet has spread to a
diameter of $40 \,\si{\milli\meter}$. State the dominant uncertainty
source and a factor-of-three confidence range.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The volumetric flow rate is $Q = 8 \,\si{\liter\per\minute} = 1.33
\times 10^{-4} \,\si{\meter\cubed\per\second}$. The nozzle area is
$A_{0} = \pi (0.006)^{2} = 1.13 \times 10^{-4} \,\si{\meter\squared}$.
The exit speed is $\norm{\vect{v}_{0}} = Q / A_{0} = 1.18 \,\si
{\meter\per\second}$, directed along the nozzle's axis. At the
$40 \,\si{\milli\meter}$ jet patch, the area has grown by a factor
$(40/12)^{2} \approx 11$; conservation of mass for an incompressible
jet gives a spread velocity $\norm{\vect{v}_{1}} \approx 0.11 \,\si
{\meter\per\second}$, again directed along the local jet axis.

The dominant uncertainty is the spread profile: the jet is not a
plug-flow column, and the boundary between jet and ambient air mixes
at a rate that grows with axial distance. A factor-of-three confidence
range on the local velocity at the patch, $0.04$ to $0.33 \,\si{\meter
\per\second}$, brackets the kind of estimate a domestic measurement
can support; a working engineer would either measure the velocity
directly with a small impeller anemometer or compute the spread from
a published self-similar jet profile. The estimate's value is the
algebraic structure: the same volumetric flow rate, divided by two
different cross-sectional areas, gives two velocities that differ by
the area ratio. The vector points along the jet axis at both points,
not at the same direction in space.
\end{estimation}

\subsection{The zero vector and linear independence}

Two vectors are \engterm{linearly dependent} if one is a scalar
multiple of the other; otherwise they are linearly independent. A set
of $k$ vectors in $\R^{n}$ is linearly independent if no non-trivial
linear combination $\alpha_{1}\vect{v}_{1} + \cdots + \alpha_{k}
\vect{v}_{k}$ equals the zero vector. A set of three vectors in
$\R^{3}$ is linearly independent if they do not all lie in a common
plane; geometrically, the parallelepiped they span has non-zero
volume. The reader meets this condition formally as the determinant
test in \cref{vol02:ch10}; for now we use the geometric criterion.

Linear independence governs whether a coordinate system is well-posed.
A pilot navigating with three compass bearings whose three directions
all lie in a horizontal plane has a well-posed two-dimensional
position fix, but no altitude information; the third bearing was not
linearly independent of the first two in $\R^{3}$. The error is not
in the magnitudes of the bearings but in their directions. Linear
independence is the discipline of checking, before measurement, that
the proposed reference frame can resolve the quantity that needs
resolving.

\section{Dot and cross products}\pathtag{core}

Two products of vectors recur in working engineering: the
\engterm{dot product}, which produces a scalar, and the
\engterm{cross product}, which produces a vector. They are the
algebraic embodiments of two distinct geometric questions.

\subsection{The dot product}

The dot product of $\vect{u}$ and $\vect{v}$ in $\R^{n}$ is
\[
\vect{u} \cdot \vect{v} = u_{1} v_{1} + u_{2} v_{2} + \cdots +
u_{n} v_{n}.
\]
The geometric form is $\vect{u} \cdot \vect{v} = \norm{\vect{u}}
\norm{\vect{v}} \cos\theta$, where $\theta$ is the angle between
$\vect{u}$ and $\vect{v}$ when their tails are joined. The two forms
agree because the law of cosines applied to the triangle with sides
$\vect{u}$, $\vect{v}$, and $\vect{u} - \vect{v}$ rearranges to the
componentwise expression \cite{text:strang2016}.

The dot product has three working engineering uses:

\begin{itemize}[leftmargin=*]
  \item \engterm{Projection}. The component of $\vect{v}$ along the
    direction $\unitvec{u}$ is $\vect{v} \cdot \unitvec{u}$. This is
    how an engineer reads off the load along a beam axis when the
    applied force is at an angle, or how a controller extracts the
    along-track and cross-track components of a vehicle's velocity
    relative to a desired course.
  \item \engterm{Orthogonality test}. Two non-zero vectors are
    perpendicular if and only if $\vect{u} \cdot \vect{v} = 0$. The
    test is exact: any deviation from zero is the cosine of the
    actual angle, scaled by the magnitudes.
  \item \engterm{Work and rate of work}. The work done by a constant
    force $\vect{F}$ over a displacement $\vect{d}$ is
    $W = \vect{F} \cdot \vect{d}$; the instantaneous power
    $P = \vect{F} \cdot \vect{v}$. Volumes III and V take up these
    two uses in the contexts of mechanics and energy systems.
\end{itemize}

The dot product is commutative ($\vect{u} \cdot \vect{v} = \vect{v}
\cdot \vect{u}$), distributes over addition, and is bilinear in its
two arguments. It is positive on a non-zero vector dotted with itself:
$\vect{v} \cdot \vect{v} = \norm{\vect{v}}^{2} \geq 0$, with
equality only for $\vect{v} = \vect{0}$. The Cauchy-Schwarz
inequality, $\abs{\vect{u} \cdot \vect{v}} \leq \norm{\vect{u}}
\norm{\vect{v}}$, follows immediately and is the algebraic content of
$\abs{\cos\theta} \leq 1$. The Cauchy-Schwarz inequality is the
discipline behind every bound that says one vector quantity controls
another.

\subsection{The cross product}

The cross product of $\vect{u}$ and $\vect{v}$ in $\R^{3}$ is the
unique vector $\vect{u} \times \vect{v}$ that is perpendicular to
both $\vect{u}$ and $\vect{v}$, has magnitude $\norm{\vect{u}}
\norm{\vect{v}} \sin\theta$, and points in the direction given by
the right-hand rule when $\vect{u}$ and $\vect{v}$ are not parallel.
In Cartesian components,
\[
\vect{u} \times \vect{v} =
\begin{pmatrix}
u_{2} v_{3} - u_{3} v_{2} \\
u_{3} v_{1} - u_{1} v_{3} \\
u_{1} v_{2} - u_{2} v_{1}
\end{pmatrix}.
\]
The componentwise expression is the determinant
\[
\vect{u} \times \vect{v} = \det
\begin{pmatrix}
\unitvec{x} & \unitvec{y} & \unitvec{z} \\
u_{1} & u_{2} & u_{3} \\
v_{1} & v_{2} & v_{3}
\end{pmatrix},
\]
when the determinant is expanded along the first row.

The cross product has three working engineering uses:

\begin{itemize}[leftmargin=*]
  \item \engterm{Area of a parallelogram}. The magnitude
    $\norm{\vect{u} \times \vect{v}} = \norm{\vect{u}}\norm{\vect{v}}
    \sin\theta$ is the area of the parallelogram with sides $\vect{u}$
    and $\vect{v}$. Triangle area is half of that. The cross product
    is the algebraic engine behind every triangulated surface area
    calculation in the second half of this chapter.
  \item \engterm{Torque and angular velocity}. The torque about a
    point produced by a force $\vect{F}$ applied at position
    $\vect{r}$ is $\vect{\tau} = \vect{r} \times \vect{F}$. The
    velocity of a point at position $\vect{r}$ on a rigid body
    rotating with angular velocity $\vect{\omega}$ is $\vect{v} =
    \vect{\omega} \times \vect{r}$. Volume III takes up rigid-body
    dynamics with these two formulas as the working tools.
  \item \engterm{Surface normal}. Given two non-parallel tangent
    vectors $\vect{t}_{u}$ and $\vect{t}_{v}$ to a surface at a
    point, the unit normal is $\unitvec{n} = (\vect{t}_{u} \times
    \vect{t}_{v}) / \norm{\vect{t}_{u} \times \vect{t}_{v}}$. The
    direction of the normal is fixed by the order of the cross
    product, which is the chapter's failure section.
\end{itemize}

The cross product is anti-commutative, $\vect{u} \times \vect{v} =
-\vect{v} \times \vect{u}$, distributes over addition, and is bilinear.
It is not associative: $\vect{u} \times (\vect{v} \times \vect{w})
\neq (\vect{u} \times \vect{v}) \times \vect{w}$ in general. The
\engterm{BAC-CAB rule}, $\vect{u} \times (\vect{v} \times \vect{w}) =
\vect{v}(\vect{u} \cdot \vect{w}) - \vect{w}(\vect{u} \cdot \vect{v})$,
expands the triple cross product into a sum of vectors weighted by
dot products, and is the working identity for unwinding a triple
cross product in a derivation.

\subsection{The scalar triple product and parallelepiped volume}

The \engterm{scalar triple product} $\vect{u} \cdot (\vect{v} \times
\vect{w})$ is a scalar whose absolute value equals the volume of the
parallelepiped spanned by $\vect{u}$, $\vect{v}$, $\vect{w}$. The
sign reports the handedness of the triple: positive when
$(\vect{u}, \vect{v}, \vect{w})$ is right-handed (in the same sense
as $\unitvec{x}, \unitvec{y}, \unitvec{z}$), negative when
left-handed. The scalar triple product equals the determinant of the
$3 \times 3$ matrix whose rows are $\vect{u}, \vect{v}, \vect{w}$:
\[
\vect{u} \cdot (\vect{v} \times \vect{w}) = \det
\begin{pmatrix}
u_{1} & u_{2} & u_{3} \\
v_{1} & v_{2} & v_{3} \\
w_{1} & w_{2} & w_{3}
\end{pmatrix}.
\]
The chapter's project rests on the scalar triple product: the
volume of a tetrahedron with one vertex at the origin and the
opposite face spanned by three vertices is one-sixth of the absolute
value of the scalar triple product of those three position vectors.

\subsection{Identities the working engineer keeps to hand}

A short list of vector identities recurs across Volumes III through
XII. Each can be derived from the componentwise definitions; the
reader is expected to know that they exist and to look them up when
needed, rather than to memorise them.

\begin{align*}
  \vect{u} \cdot (\vect{v} \times \vect{w}) &=
    \vect{v} \cdot (\vect{w} \times \vect{u}) =
    \vect{w} \cdot (\vect{u} \times \vect{v}), \\
  \vect{u} \times (\vect{v} \times \vect{w}) &=
    \vect{v}(\vect{u} \cdot \vect{w}) - \vect{w}(\vect{u} \cdot \vect{v}), \\
  (\vect{u} \times \vect{v}) \cdot (\vect{w} \times \vect{x}) &=
    (\vect{u} \cdot \vect{w})(\vect{v} \cdot \vect{x}) -
    (\vect{u} \cdot \vect{x})(\vect{v} \cdot \vect{w}), \\
  \norm{\vect{u} \times \vect{v}}^{2} &=
    \norm{\vect{u}}^{2}\norm{\vect{v}}^{2} -
    (\vect{u} \cdot \vect{v})^{2}.
\end{align*}

The last identity is sometimes called Lagrange's identity and is the
algebraic statement of $\sin^{2}\theta + \cos^{2}\theta = 1$.

\section{Coordinate systems and transformations}\pathtag{standard}

A coordinate system is a rule that assigns three numbers to each
point in $\R^{3}$. Three systems recur in working engineering, and the
discipline is to use each one where the geometry of the problem fits
the symmetry of the system.

\subsection{Cartesian, cylindrical, spherical}

\engterm{Cartesian coordinates} $(x, y, z)$ assign a triple of signed
distances along three mutually perpendicular axes; the basis
$\unitvec{x}, \unitvec{y}, \unitvec{z}$ is constant in space. A point
at position $\vect{r}$ has $\vect{r} = x\unitvec{x} + y\unitvec{y} +
z\unitvec{z}$.

\engterm{Cylindrical coordinates} $(\rho, \phi, z)$ assign a radial
distance from the $z$-axis, an azimuthal angle in the $xy$-plane
measured counterclockwise from the $x$-axis, and a height along the
$z$-axis. The conversion is $x = \rho\cos\phi$, $y = \rho\sin\phi$,
$z = z$. The basis vectors $\unitvec{\rho}, \unitvec{\phi},
\unitvec{z}$ vary with $\phi$: at azimuth $\phi$, $\unitvec{\rho} =
\cos\phi\,\unitvec{x} + \sin\phi\,\unitvec{y}$, $\unitvec{\phi} =
-\sin\phi\,\unitvec{x} + \cos\phi\,\unitvec{y}$, and $\unitvec{z}$
remains itself.

\engterm{Spherical coordinates} $(r, \theta, \phi)$ assign a radial
distance from the origin, a polar angle from the $+z$-axis, and an
azimuthal angle in the $xy$-plane. The conversion is $x = r\sin\theta
\cos\phi$, $y = r\sin\theta\sin\phi$, $z = r\cos\theta$. The basis
$\unitvec{r}, \unitvec{\theta}, \unitvec{\phi}$ varies with both
$\theta$ and $\phi$. The convention adopted here is the
physics-and-engineering convention: $\theta$ is the polar angle (zero
at the north pole, $\pi$ at the south pole), $\phi$ is the azimuthal
angle. The mathematics convention swaps the two; a reader who consults
a primary source in another field should check the convention before
trusting a formula.

The choice of system is governed by symmetry. A flow in a circular
pipe has cylindrical symmetry; cylindrical coordinates make
$\partial / \partial \phi$ vanish for many of its fields. A radiation
problem from a point source has spherical symmetry; spherical
coordinates make $\partial / \partial \theta$ and $\partial /
\partial \phi$ vanish for the radial field. A working engineer
wastes time using a coordinate system that does not match the
problem's symmetry; the coordinate-choice step is part of the
problem setup.

\subsection{Volume and area elements}

Volume integrals and surface integrals require the appropriate
infinitesimal element in the chosen system. The forms below cover
the three cases:

\begin{itemize}[leftmargin=*]
  \item Cartesian: $\dd V = \dd x \,\dd y \,\dd z$. A surface
    perpendicular to $\unitvec{z}$ has area element $\dd A = \dd x
    \,\dd y$.
  \item Cylindrical: $\dd V = \rho \,\dd\rho \,\dd\phi \,\dd z$. A
    cylindrical surface at fixed $\rho$ has $\dd A = \rho \,\dd\phi
    \,\dd z$; a flat top or bottom at fixed $z$ has $\dd A = \rho
    \,\dd\rho \,\dd\phi$.
  \item Spherical: $\dd V = r^{2} \sin\theta \,\dd r \,\dd\theta \,
    \dd\phi$. A spherical surface at fixed $r$ has $\dd A = r^{2}
    \sin\theta \,\dd\theta \,\dd\phi$.
\end{itemize}

The factors of $\rho$ and $r^{2}\sin\theta$ are the
\engterm{Jacobian determinants} of the coordinate change from
Cartesian. Volume~II Chapter~10 takes up the Jacobian as the formal
volume-distortion factor of any change of variables; here it is
enough to know the three forms above by inspection.

\subsection{Rotations as orthogonal transformations}

A \engterm{rotation} of $\R^{3}$ is a linear map that preserves
lengths and orientations. Its matrix representation $\mat{R}$ is a
$3 \times 3$ matrix satisfying $\mat{R}\transpose\mat{R} = \mat{I}$
and $\det\mat{R} = +1$. The first condition says columns are
orthonormal; the second excludes reflections, which would have
$\det\mat{R} = -1$. Volume~I Chapter~3 introduced rotations as
length-preserving transformations of position; here we generalise:
any vector quantity transforms under a rotation by left-multiplication
of its column representation by $\mat{R}$.

The composition of two rotations is again a rotation: if $\mat{R}_{1}$
and $\mat{R}_{2}$ are rotation matrices, $\mat{R}_{2} \mat{R}_{1}$ is
also a rotation matrix. The order matters, because matrix
multiplication does not commute in general; rotating first about
$\unitvec{x}$ and then about $\unitvec{y}$ produces a different
orientation from rotating first about $\unitvec{y}$ and then about
$\unitvec{x}$. The reader who has not seen this should rotate a
small block on a desk in the two orders and confirm by inspection.

A rotation about the unit axis $\unitvec{n}$ by angle $\alpha$ has
matrix
\[
\mat{R}(\unitvec{n}, \alpha) = \cos\alpha \,\mat{I} + (1 - \cos\alpha)
\,\unitvec{n}\unitvec{n}\transpose + \sin\alpha \,[\unitvec{n}]_{\times},
\]
where $[\unitvec{n}]_{\times}$ is the $3 \times 3$ matrix that, when
multiplied on a vector $\vect{v}$, returns $\unitvec{n} \times
\vect{v}$. This is Rodrigues's formula, and it is the working tool
behind every quaternion-based attitude system in aerospace, robotics,
and computer graphics.

\subsection{Active and passive transformations}

A subtle distinction recurs across the engineering record. An
\engterm{active transformation} rotates the vector while the
coordinate frame stays still; the same physical body has new
components. A \engterm{passive transformation} rotates the coordinate
frame while the vector stays still; the same physical vector has new
components. The two operations are inverses of each other: an
active rotation by $\alpha$ corresponds to a passive rotation by
$-\alpha$. A working engineer who records that a frame was rotated
when in fact a body was rotated has signed the wrong calculation.
The discipline is to state, with each rotation matrix, whether the
matrix transforms vectors or transforms frames. Volumes III and VIII
take up the active-passive distinction in the contexts of rigid-body
dynamics and inertial navigation.

\section{Curves: parametrisation, arc length, curvature}\pathtag{standard}

A \engterm{curve} in $\R^{3}$ is the image of a continuous map
$\vect{r}: [a, b] \to \R^{3}$, $t \mapsto \vect{r}(t) =
(x(t), y(t), z(t))$. The variable $t$ is the \engterm{parameter}; the
choice of parametrisation is part of the engineer's modelling
decision. A curve has many parametrisations, all describing the same
geometric set of points.

\subsection{The tangent vector}

The \engterm{tangent vector} at parameter $t$ is the derivative
\[
\vect{r}\,'(t) = \dv{\vect{r}}{t} =
\left(\dv{x}{t}, \dv{y}{t}, \dv{z}{t}\right).
\]
Its direction is the direction of motion along the curve, and its
magnitude is the speed of motion under the chosen parametrisation. A
curve and its tangent picture together form the formal record of a
trajectory: the position $\vect{r}(t)$ at each $t$ and the velocity
$\vect{r}\,'(t)$ at each $t$.

\subsection{Arc length}

The \engterm{arc length} of the curve from $t = a$ to $t = b$ is
\[
s = \int_{a}^{b} \norm{\vect{r}\,'(t)} \,\dd t.
\]
The integrand is the speed under the parametrisation; the integral
accumulates infinitesimal arc-length contributions. The arc length
is a property of the curve, not the parametrisation: any
reparametrisation that traces the same set of points in the same
direction yields the same $s$.

A particularly clean parametrisation uses the arc length itself as
the parameter. Define $s(t) = \int_{a}^{t} \norm{\vect{r}\,'(\tau)}
\,\dd\tau$; if $\vect{r}\,'$ never vanishes, $s$ is a strictly
increasing function of $t$ and can be inverted to give $t(s)$. The
\engterm{arc-length parametrisation} $\vect{r}(s) = \vect{r}(t(s))$
has the property that $\norm{\vect{r}\,'(s)} = 1$ everywhere: the
tangent is a unit vector $\unitvec{T} = \vect{r}\,'(s)$. The
arc-length parametrisation is rarely available in closed form, but
its existence simplifies many derivations.

\subsection{Curvature}

The \engterm{curvature} $\kappa$ of a curve at a point is the
magnitude of the rate of change of the unit tangent with respect to
arc length:
\[
\kappa = \norm{\dv{\unitvec{T}}{s}}.
\]
For a circle of radius $R$, $\kappa = 1/R$ at every point. For a
straight line, $\kappa = 0$. For a curve given by an arbitrary
parametrisation, the working formula is
\[
\kappa = \frac{\norm{\vect{r}\,'(t) \times \vect{r}\,''(t)}}
              {\norm{\vect{r}\,'(t)}^{3}},
\]
which can be derived by the chain rule from the arc-length form.

The \engterm{principal normal} $\unitvec{N}$ at a point of non-zero
curvature is the unit vector in the direction of $\dd\unitvec{T}/
\dd s$; the \engterm{binormal} $\unitvec{B} = \unitvec{T} \times
\unitvec{N}$. The triple $(\unitvec{T}, \unitvec{N}, \unitvec{B})$
is the \engterm{Frenet-Serret frame}, an orthonormal basis attached
to the curve at each point. The frame's derivatives along the curve
satisfy the Frenet-Serret formulas, which the reader will meet in
the Volume~III treatment of rigid-body kinematics.

Curvature controls the lateral acceleration of any object travelling
along a curved path: an object moving at speed $v$ along a curve of
radius of curvature $R = 1/\kappa$ experiences a centripetal
acceleration $a_{c} = v^{2}/R = \kappa v^{2}$. A working highway
designer uses this relation to set the maximum design speed for a
banked curve given a target lateral acceleration; a working
roller-coaster designer uses the same relation in inverse to choose
the curvature for a target sensation. The relation links the
geometric curvature of the curve to the dynamic experience of
following it.

\subsection{Line integrals}

Two line integrals recur. The \engterm{scalar line integral} of a
function $f$ along a curve is
\[
\int_{C} f \,\dd s = \int_{a}^{b} f(\vect{r}(t)) \norm{\vect{r}\,'(t)}
\,\dd t,
\]
and reports a quantity-times-arc-length, such as a mass per unit
length integrated along a wire to give total mass. The \engterm{vector
line integral} of a vector field $\vect{F}$ along a curve is
\[
\int_{C} \vect{F} \cdot \dd\vect{r} = \int_{a}^{b} \vect{F}(\vect{r}(t))
\cdot \vect{r}\,'(t) \,\dd t,
\]
and reports a quantity-times-displacement-component-along-curve, such
as the work done by a force field on a particle moving along $C$.
The vector line integral is the prototype of the circulation
integral that appears in Stokes's theorem.

\section{Surfaces: parametrisation, normals, area}\pathtag{standard}

A \engterm{parametrised surface} in $\R^{3}$ is the image of a
continuous map $\vect{r}: D \to \R^{3}$, $(u, v) \mapsto
\vect{r}(u, v)$, where $D$ is a region in the $(u, v)$ parameter
plane. The domain $D$ is rectangular for simple surfaces and more
elaborate for closed surfaces such as the sphere, where the
parametrisation either covers the surface in patches or accepts a
controlled singularity at the poles.

\subsection{Tangent vectors and the surface normal}

At each interior point of $D$, the partial derivatives
$\vect{r}_{u} = \partial\vect{r}/\partial u$ and $\vect{r}_{v} =
\partial\vect{r}/\partial v$ are tangent vectors to the surface. When
they are linearly independent, the surface is called \engterm{regular}
at that point, and the normal vector is
\[
\vect{N} = \vect{r}_{u} \times \vect{r}_{v}.
\]
The unit normal is $\unitvec{n} = \vect{N} / \norm{\vect{N}}$. The
order of the cross product fixes the direction of the normal: swapping
$u$ and $v$, or swapping the order of the cross product, reverses
the direction. The reader who has chosen a parametrisation has
implicitly chosen an orientation; the chapter's failure section is
about what goes wrong when the implicit choice and the intended
choice disagree.

\subsection{Surface area}

The infinitesimal area element is $\dd A = \norm{\vect{r}_{u} \times
\vect{r}_{v}} \,\dd u \,\dd v$, and the surface area of the
parametrised surface is
\[
A = \iint_{D} \norm{\vect{r}_{u} \times \vect{r}_{v}} \,\dd u \,\dd v.
\]
The cross-product magnitude is the area of the parallelogram spanned
by the two tangent vectors at each point; the integral accumulates
those infinitesimal parallelograms across the surface. The formula
generalises the cross-product-as-parallelogram-area identity from
section~8.2 to a continuous family of parallelograms.

\subsection{Surface integrals}

The \engterm{scalar surface integral} of a function $f$ over a
parametrised surface $S$ is
\[
\iint_{S} f \,\dd A = \iint_{D} f(\vect{r}(u, v))
\norm{\vect{r}_{u} \times \vect{r}_{v}} \,\dd u \,\dd v,
\]
and reports a quantity-times-area, such as the surface mass of a
membrane with variable density. The \engterm{flux integral} of a
vector field $\vect{F}$ through an oriented surface $S$ is
\[
\iint_{S} \vect{F} \cdot \unitvec{n} \,\dd A =
\iint_{D} \vect{F}(\vect{r}(u, v)) \cdot
(\vect{r}_{u} \times \vect{r}_{v}) \,\dd u \,\dd v,
\]
and reports the rate of transport of $\vect{F}$ across $S$ in the
direction of $\unitvec{n}$. The flux integral is the working object
of the divergence theorem and the chapter's project.

\subsection{Closed surfaces and outward normals}

A \engterm{closed surface} bounds a three-dimensional region. The
working convention is that the unit normal $\unitvec{n}$ on a closed
surface points outward, away from the bounded region. With this
convention, $\oint_{\partial V} \vect{F} \cdot \unitvec{n} \,\dd A$ is
the net outward flux of $\vect{F}$ through the boundary of the
volume $V$. A reader using a parametrisation whose induced normal
points inward must either reverse the parametrisation or change the
sign of the result. The discipline is to verify the orientation
explicitly, and the failure section returns to the operational test
that catches the error.

\section{Gradient, divergence, curl}\pathtag{core}

Three differential operators on fields recur in working engineering.
Each takes a field defined on a region of $\R^{3}$ and returns
another field. Each has an interpretation as a local rate of
something the engineer is tracking.

\subsection{The gradient}

A \engterm{scalar field} $\Phi(\vect{r})$ assigns a number to each
point. The \engterm{gradient} $\grad\Phi$ is the vector field whose
components in Cartesian coordinates are
\[
\grad\Phi = \left(\pdv{\Phi}{x}, \pdv{\Phi}{y}, \pdv{\Phi}{z}\right).
\]
The gradient at a point has two interpretations.

First, it is the direction of steepest increase of $\Phi$ at the
point. A reader walking on a heated plate with temperature field
$T(\vect{r})$ feels the temperature increase fastest in the direction
$\grad T$ and decrease fastest in the direction $-\grad T$. The
magnitude $\norm{\grad T}$ is the rate of increase per unit distance
in that direction.

Second, the gradient is the local linear approximation to the field's
variation: for small $\dd\vect{r}$,
\[
\dd\Phi \approx \grad\Phi \cdot \dd\vect{r}.
\]
This is the multivariable analogue of the scalar derivative
relation $\dd f \approx f'(x) \dd x$. The right-hand side is the
\engterm{directional derivative} of $\Phi$ in the direction
$\dd\vect{r}$.

A vector field that is the gradient of some scalar field is called
\engterm{conservative}. Conservative fields have the property that
their line integral between two endpoints does not depend on the
path connecting them: $\int_{C} \grad\Phi \cdot \dd\vect{r} =
\Phi(\vect{r}_{\text{end}}) - \Phi(\vect{r}_{\text{start}})$. Volume~III
takes up conservative force fields and the closely related notion of
potential energy.

\begin{archetype}[Transport, gradient form]
A transport quantity moves down a gradient. Heat flows from hot to
cold, with flux density $\vect{q} = -k \grad T$ for thermal
conductivity $k$. Mass flows down a concentration gradient, with
flux density $\vect{J} = -D\grad c$ for diffusivity $D$. Charge in
a conductor flows down a voltage gradient, with current density
$\vect{j} = -\sigma_{e}\grad V$ for electrical conductivity
$\sigma_{e}$. The minus sign reports that flux opposes the gradient:
the quantity moves in the direction of decreasing $\Phi$, which is
the direction $-\grad\Phi$. Volumes III, IV, and V take up the
specific cases; the archetype is the same.
\end{archetype}

\subsection{The divergence}

A \engterm{vector field} $\vect{F}(\vect{r})$ assigns a vector to each
point. The \engterm{divergence} $\divg\vect{F}$ is the scalar field
whose value in Cartesian coordinates is
\[
\divg\vect{F} = \pdv{F_{x}}{x} + \pdv{F_{y}}{y} + \pdv{F_{z}}{z}.
\]
The divergence at a point has the interpretation of net outward flux
per unit volume. Imagine a small box centred at the point; the rate
at which $\vect{F}$ flows out of the box, divided by the volume of
the box, is $\divg\vect{F}$ in the limit as the box shrinks to the
point. A positive divergence is a local source; a negative divergence
is a local sink; zero divergence is a point that conserves the
quantity carried by $\vect{F}$.

For an incompressible fluid, the velocity field $\vect{v}$ satisfies
$\divg\vect{v} = 0$ everywhere: no point is a source or sink of
volume. For a temperature field with a local heating element, the
heat flux $\vect{q}$ has $\divg\vect{q} = -\rho c \,\partial T /
\partial t$ at points away from the element and $\divg\vect{q}$ equal
to the local volumetric heat source at the element. The reader will
meet this balance as the heat equation in Volume~II Chapter~12.

\subsection{The curl}

The \engterm{curl} $\curl\vect{F}$ of a vector field is the vector
field whose Cartesian components are
\[
\curl\vect{F} =
\begin{pmatrix}
\partial F_{z}/\partial y - \partial F_{y}/\partial z \\
\partial F_{x}/\partial z - \partial F_{z}/\partial x \\
\partial F_{y}/\partial x - \partial F_{x}/\partial y
\end{pmatrix},
\]
or as the symbolic determinant
\[
\curl\vect{F} = \det
\begin{pmatrix}
\unitvec{x} & \unitvec{y} & \unitvec{z} \\
\partial/\partial x & \partial/\partial y & \partial/\partial z \\
F_{x} & F_{y} & F_{z}
\end{pmatrix}.
\]
The curl at a point has the interpretation of twice the local
angular velocity of an infinitesimal paddle wheel placed at the
point and free to rotate about an axis aligned with $\curl\vect{F}$.
A vector field with $\curl\vect{F} = \vect{0}$ everywhere is called
\engterm{irrotational}; a fluid flow whose velocity is irrotational
has no local rotation. A non-zero curl reports local circulation.

\subsection{Two identities the engineer keeps to hand}

Two identities link the three operators. The first,
\[
\divg(\curl\vect{F}) = 0
\quad\text{for any sufficiently smooth } \vect{F},
\]
says the curl of a vector field has zero divergence. The second,
\[
\curl(\grad\Phi) = \vect{0}
\quad\text{for any sufficiently smooth } \Phi,
\]
says the gradient of a scalar field has zero curl. The two identities
are the backbone of Helmholtz decomposition, which factors a
sufficiently nice vector field into a curl-free part (a gradient)
and a divergence-free part (a curl). The Volume~II Chapter~12
treatment of the Laplace and Poisson equations rests on these two
identities.

\subsection{Operators in cylindrical and spherical coordinates}

The componentwise expressions in cylindrical and spherical
coordinates are not the same as in Cartesian, because the basis
vectors themselves vary in space. The forms below are the working
results; the derivations are exercises in the chain rule that the
reader can do once and reread when needed.

\paragraph{Cylindrical $(\rho, \phi, z)$.}
\begin{align*}
\grad\Phi &= \pdv{\Phi}{\rho}\unitvec{\rho} +
  \frac{1}{\rho}\pdv{\Phi}{\phi}\unitvec{\phi} +
  \pdv{\Phi}{z}\unitvec{z}, \\
\divg\vect{F} &= \frac{1}{\rho}\pdv{(\rho F_{\rho})}{\rho} +
  \frac{1}{\rho}\pdv{F_{\phi}}{\phi} + \pdv{F_{z}}{z}.
\end{align*}

\paragraph{Spherical $(r, \theta, \phi)$.}
\begin{align*}
\grad\Phi &= \pdv{\Phi}{r}\unitvec{r} +
  \frac{1}{r}\pdv{\Phi}{\theta}\unitvec{\theta} +
  \frac{1}{r\sin\theta}\pdv{\Phi}{\phi}\unitvec{\phi}, \\
\divg\vect{F} &= \frac{1}{r^{2}}\pdv{(r^{2} F_{r})}{r} +
  \frac{1}{r\sin\theta}\pdv{(\sin\theta\, F_{\theta})}{\theta} +
  \frac{1}{r\sin\theta}\pdv{F_{\phi}}{\phi}.
\end{align*}

The radial factors $\rho$, $r^{2}$, and $r\sin\theta$ in the
divergence expressions are the Jacobians of the surface elements at
constant $\rho$, $r$, and $\theta$ respectively. They are the same
factors that appeared in the volume and area elements above; the
agreement is not a coincidence but a statement of the divergence
theorem at infinitesimal scale.

\section{Stokes, Gauss, Green: the three integral theorems}\pathtag{core}

The three integral theorems of vector calculus link the local
behaviour of a field to its accumulated boundary behaviour. Each is
a balance statement: an interior accumulation equals a boundary flux.
The three theorems are the chapter's formal home for the balance
archetype.

\subsection{Green's theorem}

Green's theorem connects a line integral around a simple closed curve
in the plane to a double integral over the region the curve encloses.
For a vector field $\vect{F} = (P, Q)$ in $\R^{2}$ and a simple
closed curve $C$ traversed counterclockwise, enclosing the region
$D$,
\[
\oint_{C} \vect{F} \cdot \dd\vect{r} =
\oint_{C} (P \,\dd x + Q \,\dd y) =
\iint_{D} \left(\pdv{Q}{x} - \pdv{P}{y}\right) \dd A.
\]
The left side is the circulation of $\vect{F}$ around $C$; the right
side is the integral over $D$ of the scalar that a two-dimensional
reader recognises as the $z$-component of the curl. Green's theorem
is Stokes's theorem in two dimensions.

A working application: the area of a planar region $D$ bounded by
$C$ equals
\[
\text{Area}(D) = \oint_{C} x \,\dd y = -\oint_{C} y \,\dd x =
\frac{1}{2}\oint_{C} (x \,\dd y - y \,\dd x).
\]
The third form is the antisymmetric \engterm{shoelace formula}: a
land surveyor walking the boundary of a parcel and recording $x$ and
$y$ at each turn computes the parcel's area by accumulating
$\tfrac{1}{2}(x_{i} y_{i+1} - x_{i+1} y_{i})$ around the boundary.

\subsection{The divergence theorem (Gauss)}

The \engterm{divergence theorem} connects the volume integral of the
divergence of a vector field over a region to the flux of the field
through the region's boundary. For a sufficiently smooth vector field
$\vect{F}$ on a region $V \subset \R^{3}$ with closed, oriented
boundary $\partial V$ whose unit normal points outward,
\[
\iiint_{V} \divg\vect{F} \,\dd V =
\oint_{\partial V} \vect{F} \cdot \unitvec{n} \,\dd A.
\]
The integrand on the left is the local rate of net outflow per unit
volume, integrated over the volume. The right side is the integrated
net flux through the boundary. The theorem says the two are equal:
whatever is created or destroyed inside $V$ must, in the steady
state, cross $\partial V$.

\begin{archetype}[Balance, divergence theorem form]
A balance is an accounting equation. Volume~I introduced the form
``what is inside equals what came in minus what went out, plus what
was created.'' The divergence theorem is the same statement in the
calculus of fields: the total interior source rate equals the total
boundary outflow rate. Mass conservation in fluid mechanics, charge
conservation in electromagnetism, energy conservation in
thermodynamics, and species conservation in reactive transport are
all instances of this single archetype, with $\vect{F}$ being the
appropriate flux density and $\divg\vect{F}$ the appropriate source
density.
\end{archetype}

A worked specialisation of the theorem produces the chapter's
project formula. Take $\vect{F}(\vect{r}) = \vect{r}$, the position
vector field. Its divergence in Cartesian coordinates is
\[
\divg\vect{r} = \pdv{x}{x} + \pdv{y}{y} + \pdv{z}{z} = 3.
\]
The divergence theorem then reads
\[
\iiint_{V} 3 \,\dd V = \oint_{\partial V} \vect{r} \cdot \unitvec{n}
\,\dd A,
\]
which rearranges to
\[
V = \frac{1}{3} \oint_{\partial V} \vect{r} \cdot \unitvec{n} \,\dd A.
\]
The volume of any region equals one-third of the surface integral of
$\vect{r} \cdot \unitvec{n}$ over its closed boundary. This is the
formula the project asks the reader to apply to a sampled household
object.

\subsection{Stokes's theorem}

\engterm{Stokes's theorem} connects the surface integral of the curl
of a vector field over an oriented surface to the line integral of
the field around the surface's boundary. For a sufficiently smooth
vector field $\vect{F}$ on a surface $S$ with positively oriented
boundary $\partial S$,
\[
\iint_{S} (\curl\vect{F}) \cdot \unitvec{n} \,\dd A =
\oint_{\partial S} \vect{F} \cdot \dd\vect{r}.
\]
The orientation convention is the right-hand rule: with the right
hand wrapped around the boundary in the direction of $\dd\vect{r}$,
the thumb points along $\unitvec{n}$. The left side is the total
``twist'' of $\vect{F}$ over $S$; the right side is the
circulation of $\vect{F}$ around the boundary $\partial S$. The
theorem says the two are equal: a surface integral of curl reports
the same number as a line integral around the surface's edge.

The working consequence is that a circulation depends only on the
curl of the field through any surface spanning the curve, not on the
particular surface chosen. Two different surfaces with the same
boundary give the same circulation. The reader meets this
path-independence again in Volume~III, where it is the working
basis of magnetostatics, and in Volume~IV, where it is the
working basis of electromagnetism.

\subsection{Why three theorems and not one}

The three theorems are three faces of the same fundamental statement.
The general form, the \engterm{generalised Stokes theorem}, is
\[
\int_{M} \dd\omega = \int_{\partial M} \omega,
\]
for a sufficiently nice differential form $\omega$ on a sufficiently
nice manifold $M$ \cite{text:strang2016}. Green is the case $M$ is a
region in $\R^{2}$ and $\omega$ is a 1-form. Gauss is the case $M$ is
a region in $\R^{3}$ and $\omega$ is a 2-form. Stokes is the case
$M$ is a surface in $\R^{3}$ and $\omega$ is a 1-form. The reader
who studies differential forms in a later course will recognise the
three theorems as the same theorem in three dimensions and codimensions;
the working engineer who does not study forms still has the three
working tools, with the same orientation discipline, the same
boundary discipline, and the same balance interpretation.

\subsection{The working consequence: balance is a theorem}

\begin{principle}[Balance is a theorem]
A balance equation, written for any quantity that admits a flux
density and a source density, is the divergence theorem applied to
that quantity. The interior accumulation rate is the volume integral
of the source density. The boundary flux rate is the surface integral
of the flux density dotted with the outward normal. The two are
equal by the divergence theorem. A reader who has internalised this
fact has reduced ten domain-specific conservation laws to one
mathematical identity applied ten times.
\end{principle}

The principle is what the chapter offers Volume~III and beyond. The
mass-conservation equation of fluid mechanics, the charge-conservation
equation of electromagnetism, the energy-conservation equation of
thermodynamics, the species-conservation equation of chemical
engineering, and the population-conservation equation of biological
modelling are all the same identity. Each volume restates it in the
domain's working notation; the reader who has met the identity here
recognises the restatement as a notational variant rather than as a
new principle.

\section{Failure: orientation errors and the sign-of-the-normal
problem}\pathtag{core}

Vector calculus has one failure mode that recurs in working
engineering more often than any other: the orientation error. The
sign of a flux integral, the direction of a torque, the chirality of
a coordinate frame, and the inward-versus-outward convention on a
closed surface all depend on a binary choice. A working engineer who
gets the sign wrong has produced a calculation that looks correct in
its magnitudes and is reversed in its direction; a working engineer
who never checks the sign explicitly is one orientation error away
from a sign-reversal in a load, a flow, or a control output.

\subsection{The mechanism}

The orientation error has three common forms.

\engterm{The hand-rule swap}. Cross products and curl results depend
on the right-hand rule. A reader who applies the left-hand rule, or
who applies the right-hand rule with a left-handed coordinate frame,
produces a vector that points in the opposite direction. The error
propagates: a torque computed with the wrong hand rule reverses the
direction of the angular acceleration; a magnetic-field direction
computed with the wrong hand rule reverses the force on a current.
Some engineering communities use left-handed coordinate frames as a
local convention (the screen-coordinate frame in many graphics
systems is left-handed); a working engineer crossing between
communities checks the convention before trusting the formula.

\engterm{The parametrisation flip}. A surface parametrisation
$\vect{r}(u, v)$ implies a normal direction $\unitvec{n} =
(\vect{r}_{u} \times \vect{r}_{v}) / \norm{\vect{r}_{u} \times
\vect{r}_{v}}$. Swapping the roles of $u$ and $v$, or reversing the
orientation of either parameter, reverses $\unitvec{n}$. A flux
integral computed with the reversed normal is the negative of the
intended one. The reader who chooses a parametrisation without
checking which direction the normal points has implicitly chosen a
sign for every flux integral over the surface.

\engterm{The closed-surface orientation lapse}. The divergence
theorem requires the unit normal on a closed surface to point
outward. A reader who parametrises a closed surface piecewise
(top, bottom, sides) and uses the same parametrisation rule for
each piece often discovers, midway through the surface integral,
that one piece's normal points inward. The flux contributions of
the pieces with reversed normals are subtracted instead of added,
and the result is wrong by an unknown amount that depends on the
proportion of misoriented pieces.

\subsection{The operational test}

The discipline that catches all three forms is the same: an explicit
operational test of the orientation, performed before any
quantitative calculation, and rerun whenever the parametrisation
changes.

For a closed surface, the test is a sample point. Choose a point on
the surface, evaluate $\unitvec{n}$ at that point, and check by
inspection whether the resulting vector points away from the
enclosed volume. If it does, the orientation is outward. If it does
not, swap the parametrisation order or negate $\unitvec{n}$ before
proceeding.

For a curl-and-circulation calculation, the test is a sample
direction. Trace the boundary curve in the parametrised direction;
align the right hand so the fingers curl with the boundary. The
thumb's direction is the surface normal as Stokes's theorem demands
it. If the chosen $\unitvec{n}$ disagrees, reverse the boundary
parametrisation or negate the normal.

For a torque or angular-velocity calculation, the test is a sample
load. Apply a positive force along $\vect{F}$ at position $\vect{r}$;
imagine which way the system actually spins. The torque vector
$\vect{r} \times \vect{F}$ should point along the axis whose
right-hand rotation matches the imagined spin. If the formal
computation produces the opposite axis, the cross-product order or
the sign convention is wrong.

\subsection{The rule the engineer applies}

\begin{principle}[Verify orientation explicitly]
Every flux integral, surface integral, line integral, cross product,
curl, and torque carries an implicit sign that depends on a binary
orientation choice. The engineer states the choice explicitly,
verifies it against an operational sample at the time the choice is
made, and rechecks it whenever the parametrisation, the basis, or the
hand rule changes. A calculation whose magnitudes are correct and
whose sign is wrong has the same effect as a calculation that is
wrong by a factor of two: the error is loud, traceable, and
unforgivable in a system that depends on the sign for safety,
direction, or stability.
\end{principle}

The principle has a corollary. A reader who is asked to inherit a
calculation from a previous engineer should rerun the orientation
test before extending the calculation. The previous engineer's
parametrisation is the previous engineer's responsibility; the
calculation about to be performed is the present engineer's
responsibility, and the orientation discipline does not transfer
across the inheritance. Volumes III, IV, and IX take up specific
cases of inherited-orientation failure in mechanics, electromagnetism,
and aerospace; the rule above is the general one.

\subsection{A worked example of the failure}

A reader computes the flux of $\vect{F} = (0, 0, z)$ outward through
a closed surface composed of the unit cube's six faces, by
parametrising each face as a function of two of the Cartesian
coordinates with a fixed value of the third. On the top face,
$z = 1$, the reader writes $\vect{r}(x, y) = (x, y, 1)$ for $(x, y)
\in [0, 1]^{2}$. The induced normal is $\vect{r}_{x} \times
\vect{r}_{y} = \unitvec{x} \times \unitvec{y} = +\unitvec{z}$, which
is outward. On the bottom face, $z = 0$, the reader writes
$\vect{r}(x, y) = (x, y, 0)$ with the same parametrisation rule. The
induced normal is again $+\unitvec{z}$, but on the bottom face
$+\unitvec{z}$ points inward, not outward.

A reader who carries this orientation through the surface integral
adds the bottom face's flux $\iint_{\text{bottom}} \vect{F} \cdot
(+\unitvec{z}) \,\dd A = 0$ instead of subtracting it. In this
specific case, $\vect{F} = (0, 0, z)$ vanishes on the bottom face,
and the error of sign produces no error in magnitude. A different
field, $\vect{F} = (0, 0, 1)$, exposes the failure: each side face
contributes zero, the top contributes $+1$ (correctly), and the
bottom contributes $+1$ in the reader's convention versus the
correct $-1$. The reader's result is $+2$; the correct result is $0$.
The divergence theorem confirms: $\divg\vect{F} = 0$, so
$\iiint_{V} \divg\vect{F} \,\dd V = 0$.

The discipline this example asks of the reader is mechanical. Before
the surface integral is computed, evaluate $\unitvec{n}$ on each face
at a sample point and confirm it points outward. The five-second
check at parametrisation time prevents the half-hour debugging hunt
at result-checking time, and prevents the much longer professional
embarrassment when a result reaches a colleague before the sign
error is caught.

\section{Project}\pathtag{core}

\begin{project}[The volume of an irregular object, by the divergence
theorem]
The reader takes one irregular household object whose volume the
reader already measured by water displacement in the Volume~I
Chapter~7 project, and computes that volume a second way: by the
divergence theorem applied to a sampled triangulation of the object's
surface. The two values are compared. Where they agree within their
combined uncertainty, the divergence theorem and the empirical
measurement reinforce each other; where they disagree, the reader
diagnoses whether the surface sample is too coarse, the divergence
formula is being applied with a sign error, or the water-displacement
measurement was biased. The chapter has just finished arguing that
balance is a theorem; the project is the reader's first occasion to
exercise that theorem on a measured object.

\textbf{Track}. Household and analysis (hybrid). The reader works at
the kitchen table or workbench. \textbf{Hazard class}: none. The
chosen object should not be sharp, electrical, valuable, or
water-damaged. \textbf{Tools}. The same household tools as
Volume~I Chapter~7 (kitchen scale, ruler, beaker or graduated
cylinder), plus paper and pencil for the cross-section sampling and
either a programming environment (Python, MATLAB, Julia, or
equivalent) or a spreadsheet for the surface-mesh triangulation and
the divergence-formula calculation. A 3D scanner or smartphone
photogrammetry app is optional and useful but not required.

\textbf{Choosing the object}. Reuse the irregular object measured in
Volume~I Chapter~7 if possible: the comparison against the original
water-displacement value is the project's reconciliation step. If
the original object is no longer available, choose a comparable
object (a smooth stone, a small piece of fruit, a small hand tool)
and rerun the water-displacement measurement first.

\textbf{Procedure}. Apply the formula
\[
V = \frac{1}{3} \oint_{\partial V} \vect{r} \cdot \unitvec{n} \,\dd A
\]
to a sampled approximation of the object's surface, in five steps.

\begin{enumerate}[leftmargin=*]
  \item \engterm{Establish a coordinate frame}. Place the object on
    a sheet of paper in a stable orientation. Mark a chosen origin
    on the paper directly under a chosen reference point on the
    object (a stable resting point or a marked feature). Define the
    $z$-axis pointing up; choose two perpendicular axes in the paper
    plane and label them $x$ and $y$. Verify the frame is right-
    handed: $\unitvec{x} \times \unitvec{y}$ should point along
    $+\unitvec{z}$.

  \item \engterm{Sample at least 30 surface points}. Use a calliper,
    ruler, or other measuring tool to record the $(x, y, z)$
    coordinates of at least 30 points distributed across the
    object's surface. Distribute the points so that no part of the
    surface is unsampled by more than approximately one-third of the
    object's smallest characteristic dimension. The points should be
    organised either by horizontal cross-sections at several heights
    or by latitude-and-longitude rings around the object. Record the
    sampling resolution as part of the measurement record.

  \item \engterm{Build a closed triangulated surface}. Connect the
    sampled points into a closed triangle mesh that approximates the
    object's surface. The simplest construction stacks the
    horizontal cross-sections, connects each to the next by a band
    of triangles, and caps the top and bottom with fan triangulations
    centred on the highest and lowest sampled points. Each triangle
    is specified by an ordered triple of vertex labels; the order
    determines the triangle's normal direction. Use the same
    orientation convention for every triangle (for instance,
    counterclockwise as seen from outside the object), so the
    resulting mesh has consistently outward-pointing normals.
    Verify the convention by sampling one triangle and checking by
    inspection that the right-hand rule applied to its vertex order
    gives a normal that points away from the enclosed volume.

  \item \engterm{Compute the divergence-theorem volume}. For each
    triangle with ordered vertices $\vect{p}_{1}, \vect{p}_{2},
    \vect{p}_{3}$, the contribution to $V$ is
    \[
    V_{\triangle} = \frac{1}{6}
    \vect{p}_{1} \cdot (\vect{p}_{2} \times \vect{p}_{3}).
    \]
    This is the signed volume of the tetrahedron with one vertex at
    the origin and the opposite face the triangle. The total volume
    is the sum of these contributions over all triangles:
    $V = \sum_{\triangle} V_{\triangle}$. The derivation is in the
    chapter's worked example above; the formula follows from the
    divergence-theorem expression $V = \tfrac{1}{3} \oint
    \vect{r} \cdot \unitvec{n} \,\dd A$ specialised to a flat
    triangle, where the triangle's contribution to the integral is
    $\tfrac{1}{3}\vect{r}_{\text{centroid}} \cdot \unitvec{n} \,
    A_{\triangle}$ and a short calculation reduces it to the
    one-sixth scalar triple product above. Compute the sum.

  \item \engterm{Reconcile against water displacement}. Compare the
    divergence-theorem volume $V_{\text{DT}}$ to the
    water-displacement volume $V_{\text{WD}}$ from Volume~I
    Chapter~7, with both reported with their standard uncertainties.
    Compute the compatibility ratio
    \[
    r = \frac{\abs{V_{\text{DT}} - V_{\text{WD}}}}
             {\sqrt{u(V_{\text{DT}})^{2} + u(V_{\text{WD}})^{2}}}
    \]
    using the chapter~7 cross-checking discipline. A value of $r$
    near $1$ or below means the two methods agree within their
    combined uncertainty.
\end{enumerate}

\textbf{Uncertainty budget for the divergence-theorem volume}. The
divergence-theorem volume inherits uncertainty from three sources.
The first is the position uncertainty on each sampled point, which
propagates through the scalar triple product to a per-triangle
volume uncertainty; the second is the truncation error of replacing
a smooth surface by a piecewise-flat one, which decreases as the
sample is refined; the third is the orientation discipline, which
either contributes zero error if the convention is consistent or
contributes a sign-flipped triangle's worth of error if it is not.
The reader reports each source separately and combines them in
quadrature for an expanded uncertainty at $k = 2$.

\textbf{Reflection (1500 words)}. The reader writes a 1500-word
narrative addressing:

\begin{itemize}[leftmargin=*]
  \item How the divergence-theorem volume compared to the
    Volume~I water-displacement volume, both in magnitude and in
    the compatibility ratio.
  \item Which of the three uncertainty sources dominated the
    divergence-theorem result, and what step would most reduce the
    dominant source.
  \item Whether refining the surface sampling from $30$ points
    to a finer grid would close any remaining gap, or whether the
    remaining disagreement points to a different error source.
  \item One physical engineering question (mass of a casting,
    drag on a body, heat loss from a surface) where the same
    surface-to-volume bookkeeping would be useful, and whether the
    divergence-theorem method would be the working tool or a
    pedagogical aid.
  \item What the divergence theorem buys over the empirical
    water-displacement method: in particular, where the theorem
    earns its place as a working tool and where water displacement
    is the cheaper and more defensible measurement.
\end{itemize}

\textbf{Deliverable}. The cross-section data, the triangulation, the
per-triangle volume contributions, the total divergence-theorem
volume with its uncertainty budget, the compatibility ratio against
Volume~I, and the reflection. The general editor's reference set of
divergence-theorem computations for a representative stone is
published online; the reader is encouraged to compare \emph{after}
completing their own.

\textbf{Time}. Two to four hours for sampling and triangulation, one
to two hours for the divergence calculation and uncertainty budget,
two to three hours for the reflection. Total approximately five to
nine hours, distributed across one or two days.
\end{project}

\section{Exercises}

The exercises test calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment. The editor
should attach solutions and rubrics before release.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Two vectors are $\vect{u} = (3, -1, 2)$ and $\vect{v} = (1, 4, -2)$.
Compute $\vect{u} + \vect{v}$, $\vect{u} - \vect{v}$,
$\norm{\vect{u}}$, $\norm{\vect{v}}$, and the unit vector
$\unitvec{u}$.
\end{exercise}

\begin{exercise}
For the vectors of the previous exercise, compute $\vect{u} \cdot
\vect{v}$, $\vect{u} \times \vect{v}$, the angle $\theta$ between
them, and the area of the parallelogram they span.
\end{exercise}

\begin{exercise}
Three vectors are $\vect{u} = (1, 2, 0)$, $\vect{v} = (0, 1, 3)$,
$\vect{w} = (2, 0, 1)$. Compute the scalar triple product
$\vect{u} \cdot (\vect{v} \times \vect{w})$ and report the volume
of the parallelepiped they span.
\end{exercise}

\begin{exercise}
A point at Cartesian coordinates $(2.0, 2.0, 3.0) \,\si{\meter}$ is
expressed in cylindrical coordinates $(\rho, \phi, z)$ and in
spherical coordinates $(r, \theta, \phi)$. Compute both
representations to three significant figures.
\end{exercise}

\begin{exercise}
A scalar field $\Phi(x, y, z) = x^{2} y + y z^{2}$ has a gradient
$\grad\Phi$ at every point. Compute $\grad\Phi$ at the point
$(1, 2, -1)$ and the directional derivative of $\Phi$ at that
point in the direction of the unit vector $\unitvec{u} =
(1, 1, 1)/\sqrt{3}$.
\end{exercise}

\begin{exercise}
A vector field is $\vect{F}(x, y, z) = (xy, y^{2} z, x z^{2})$.
Compute $\divg\vect{F}$ and $\curl\vect{F}$ at the point
$(1, 1, 1)$.
\end{exercise}

\begin{exercise}
A curve is parametrised by $\vect{r}(t) = (\cos t, \sin t,
t)$ for $t \in [0, 2\pi]$. Compute the tangent vector
$\vect{r}\,'(t)$, its magnitude, the arc length of the curve, and
the curvature at $t = \pi/2$.
\end{exercise}

\begin{exercise}
A surface is parametrised by $\vect{r}(u, v) = (u\cos v, u\sin v,
u^{2})$ for $u \in [0, 1]$ and $v \in [0, 2\pi]$. Compute the
tangent vectors $\vect{r}_{u}$, $\vect{r}_{v}$, the normal vector
$\vect{N} = \vect{r}_{u} \times \vect{r}_{v}$ at $(u, v) = (1/2,
\pi/4)$, and the surface area.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Starting from the componentwise definition of the dot product, derive
the geometric formula $\vect{u} \cdot \vect{v} = \norm{\vect{u}}
\norm{\vect{v}} \cos\theta$ using the law of cosines applied to the
triangle with sides $\vect{u}$, $\vect{v}$, and $\vect{u} - \vect{v}$.
\end{exercise}

\begin{exercise}
Derive Lagrange's identity $\norm{\vect{u} \times \vect{v}}^{2} =
\norm{\vect{u}}^{2}\norm{\vect{v}}^{2} - (\vect{u} \cdot
\vect{v})^{2}$ by expanding the components of the cross product and
collecting terms. State the identity's interpretation as the
algebraic content of $\sin^{2}\theta + \cos^{2}\theta = 1$.
\end{exercise}

\begin{exercise}
Derive the BAC-CAB rule, $\vect{u} \times (\vect{v} \times \vect{w})
= \vect{v}(\vect{u} \cdot \vect{w}) - \vect{w}(\vect{u} \cdot
\vect{v})$, by expanding both sides in Cartesian components and
showing they are equal term by term.
\end{exercise}

\begin{exercise}
Derive the divergence-theorem formula for the volume of a region,
$V = \tfrac{1}{3} \oint \vect{r} \cdot \unitvec{n} \,\dd A$, by
applying the divergence theorem to the position vector field
$\vect{F}(\vect{r}) = \vect{r}$. State the regularity assumption on
the boundary that the derivation requires.
\end{exercise}

\begin{exercise}
Derive the formula $V_{\triangle} = \tfrac{1}{6}\vect{p}_{1} \cdot
(\vect{p}_{2} \times \vect{p}_{3})$ for the contribution of a single
oriented triangle with vertices $\vect{p}_{1}, \vect{p}_{2},
\vect{p}_{3}$ to the divergence-theorem volume of a closed
triangulated surface. Start from the divergence-theorem expression
$V = \tfrac{1}{3} \oint \vect{r} \cdot \unitvec{n} \,\dd A$ and
specialise it to a flat triangle.
\end{exercise}

\begin{exercise}
Derive the cylindrical-coordinate expression for $\divg\vect{F}$ by
applying the divergence theorem to a small cylindrical-shell volume
element of dimensions $\dd\rho \times \rho\,\dd\phi \times \dd z$
and computing the net flux through its six faces in the limit as the
volume shrinks to a point.
\end{exercise}

\begin{exercise}
Show that the curl of any gradient vanishes: $\curl(\grad\Phi) =
\vect{0}$ for every twice-continuously-differentiable scalar field
$\Phi$. State the assumption on $\Phi$ that the derivation uses.
\end{exercise}

\begin{exercise}
Show that the divergence of any curl vanishes: $\divg(\curl\vect{F})
= 0$ for every twice-continuously-differentiable vector field
$\vect{F}$. State the assumption on $\vect{F}$ that the derivation
uses.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the surface area of a household object of the reader's
choice (a coffee mug, an apple, a small stone) using a vertex-and-
triangle approximation with no more than 12 sampled points. State
the dominant uncertainty source and a factor-of-three confidence
range. Compare to a more careful measurement if one is available.
\end{exercise}

\begin{exercise}
Estimate the volume of a household object using the divergence-
theorem formula $V = \tfrac{1}{3}\oint \vect{r}\cdot\unitvec{n}
\,\dd A$ on a coarse 8-point sampled approximation. Compare to the
estimate one would obtain from a single overall length-times-width-
times-height bounding box, and identify which estimate is more
defensible and why.
\end{exercise}

\begin{exercise}
Estimate the magnitude of the gradient of the temperature field
inside a household oven during the first 60 seconds after the door
is closed, in $\,\si{\kelvin\per\meter}$. State the assumptions
about the heat source location, the oven's interior dimensions, and
the temperature difference between the heater and the air.
\end{exercise}

\begin{exercise}
Estimate the divergence of the velocity field of room air during
the first few seconds after a household window is opened on a
breezy day. State whether the divergence is positive (air entering
the room is a volume source), negative (air leaving the room is a
volume sink), or near zero (incompressible-flow approximation), and
state the regime in which each answer is defensible.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement, in a programming environment of choice, the
divergence-theorem volume formula
$V = \tfrac{1}{6}\sum_{\triangle} \vect{p}_{1}\cdot(\vect{p}_{2}
\times \vect{p}_{3})$ for a triangulated closed surface. Test the
implementation against a known-volume shape (a unit cube
triangulated into 12 triangles, a regular tetrahedron, a sphere
triangulated into a geodesic mesh of at least 80 triangles). Report
the relative error of each test and identify the regime in which
the truncation error of the piecewise-flat approximation
dominates the position-sampling uncertainty.
\end{exercise}

\begin{exercise}
Implement a Monte Carlo estimate of the surface area of a
parametrised surface $\vect{r}(u, v) = (u\cos v, u\sin v, u^{2})$,
$u \in [0, 1]$, $v \in [0, 2\pi]$. Use at least $10\,000$
uniformly distributed sample points in $(u, v)$, evaluate the
local area density $\norm{\vect{r}_{u} \times \vect{r}_{v}}$ at
each, and report the mean and standard deviation of the area
estimate. Compare against the analytical integral.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a sampling strategy for measuring the volume of an irregular
metal casting by the divergence theorem to within $1\,\%$ relative
uncertainty. Specify the number of surface sample points, the
expected position uncertainty per point, and the truncation
uncertainty of the piecewise-flat surface approximation. State
which uncertainty source dominates at the chosen sampling density
and which would dominate at twice or half the density.
\end{exercise}

\begin{exercise}
Design a procedure for verifying the orientation convention of a
triangulated closed surface mesh produced by an unspecified
software tool. State the operational test, the expected output for
a correctly oriented mesh, and the corrective action when the test
indicates inverted normals on some triangles.
\end{exercise}

\begin{exercise}
Design a coordinate frame for a fluid-flow measurement in a
circular pipe of inner radius $50 \,\si{\milli\meter}$. State which
of Cartesian, cylindrical, or spherical coordinates is the working
choice; identify the symmetry that selects the choice; and write
the divergence operator $\divg\vect{v}$ for the velocity field in
the chosen coordinates.
\end{exercise}

\begin{exercise}
Design a procedure for computing the area of a planar parcel of
land from a surveyed sequence of $(x, y)$ corner coordinates using
the shoelace formula. State the surveying tolerance required for
the area to be known to within $1 \,\si{\meter\squared}$ on a
parcel of nominal area $1000 \,\si{\meter\squared}$.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{standard}

\begin{exercise}
A divergence-theorem volume calculation on a triangulated
household-object surface returns a value $V_{\text{DT}}$ that is
$15\,\%$ smaller than the water-displacement volume $V_{\text{WD}}$
measured for the same object. State three plausible mechanisms (the
sampling is too coarse, one or more triangles has an inverted
normal, the water-displacement reading included surface bubbles)
and propose, for each, the test that would confirm or refute it.
\end{exercise}

\begin{exercise}
A finite-element heat-conduction code reports a positive divergence
of the heat flux at every interior point of a steel block held at
constant temperature. State whether the report is physically
plausible and identify the most likely software-side mechanism if it
is not.
\end{exercise}

\begin{exercise}
A torque computation in an inertial-navigation simulation returns a
torque vector pointing along $-\unitvec{z}$ when the corresponding
physical experiment shows the platform rotating about $+\unitvec{z}$.
State the most likely cause and the operational test that would
distinguish a hand-rule swap from a passive-versus-active
transformation error.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Construct a worked example, parallel to the cube example in
section~8.8, in which a piecewise parametrisation of a closed
surface produces inverted normals on at least one face, and in
which the resulting flux integral differs from the divergence-
theorem volume integral by a known amount. State the operational
test that would catch the error before the calculation is
completed.
\end{exercise}

\begin{exercise}
A team computes the volume of a triangulated closed surface mesh
using $V = \tfrac{1}{6}\sum_{\triangle} \vect{p}_{1} \cdot
(\vect{p}_{2} \times \vect{p}_{3})$ and obtains a value smaller in
magnitude than the bounding-box volume by a factor of two and of
the wrong sign. State the most likely failure (an inverted
orientation convention applied uniformly to every triangle, so the
sum reports $-V$ instead of $+V$) and explain why the magnitude
agrees with the bounding-box estimate scaled by the
volume-fraction the bounding-box overestimates.
\end{exercise}

\begin{exercise}
A student computes the curl of a velocity field for an
incompressible fluid flow using a left-handed coordinate frame and
the right-hand rule, and is surprised when the predicted vorticity
direction disagrees with a smoke-visualisation experiment. State
the corrective action and identify the engineering domain (computer
graphics, robotics, geodesy) in which left-handed frames are common
enough that the cross-domain reader should check the convention
before trusting any cross product.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
Argue, in 200 to 300 words, whether the divergence theorem or the
empirical water-displacement method is the more defensible volume
measurement for an irregular household object. Take a position;
defend it with a specific scenario in which your preferred answer
fails and the other method is preferred.
\end{exercise}

\begin{exercise}
A senior colleague tells the reader that ``vector calculus is for
physicists; in industry we use software, and the software handles
the orientations.'' Write a one-paragraph reply that takes the
colleague seriously, identifies the regime in which the claim is
defensible, and identifies the regime in which it is not.
\end{exercise}

\begin{exercise}
A working engineer computes a quantity by both a vector-calculus
derivation and a direct numerical simulation, and the two disagree
by $5\,\%$. Argue, in one paragraph, which of the following actions
is more defensible: trust the analytical derivation and search for
a software bug; trust the simulation and search for a missed term in
the derivation; treat both as candidate answers and seek a third
independent method to break the tie. Reference the cross-checking
discipline of Volume~I Chapter~7 in defending the position.
\end{exercise}
