\chapter{Organisational failure}
\label{vol10:ch10}

\epigraph{The engineers knew. The organisation that employed the engineers decided not to know.}{}

\chapmeta{Half-life: durable. Archetypes: ethics, failure. Exercise target: 15. Project track: hybrid.}

\noindent
Many of the most consequential engineering accidents of the past century were organisational accidents in this precise sense: the technical risk was known inside the organisation, the warnings were raised inside the organisation, the responsible decision-makers had access to the information that would have prevented the accident, and the accident occurred anyway because the organisation's decision processes failed to act on what its members knew. Challenger, Columbia, Bhopal, Chernobyl, the 737 MAX, BP Texas City, Deepwater Horizon, Volkswagen Dieselgate: in each, the post-accident investigation reads as much like an organisational ethnography as like an engineering report. This chapter is about that pattern, the discipline that studies it, the mechanisms by which it occurs, and the cases that have become the field's canon.

\section{Organisations as failure-producing systems}\pathtag{core}

The classical engineering treatment of safety assumes the organisation is a neutral container for engineering work: engineers identify hazards, design controls, document procedures, train operators, and the organisation's role is to provide the resources and authority for these activities. The organisational-accident literature, beginning with Perrow's \emph{Normal Accidents} \cite{gen:perrow1984} and developed through Vaughan, Reason, Hollnagel, Dekker, and Snook, replaces the assumption with the empirical claim that organisations have characteristic failure-producing properties: incentive structures, communication patterns, authority distributions, and cultural commitments that systematically distort the engineering work the organisation contains. The organisation is not a neutral container; it is itself a component of the system whose failure modes must be designed against.

Perrow's framework distinguishes systems by their coupling (how tightly events in one part of the system propagate to others) and their interactive complexity (how many ways components can interact in unexpected sequences). High-coupling, high-complexity systems (nuclear power, chemical processing, aerospace, large software systems) are particularly prone to "normal accidents": failures that arise not from any single component's defect but from the interaction of components in ways the designers did not anticipate. The organisational structures that operate these systems shape which interactions get attention and which do not; an organisation whose attention is captured by production targets will not attend to the interaction patterns that produce the accident.

The structural patterns that produce organisational failure recur across cases and industries. The chapter sections develop the dominant ones: normalisation of deviance (where signals that should be alarming are reclassified as routine until the alarming signal is acted on); production pressure (where schedule and cost incentives erode the safety margins that engineering analysis specified); whistleblower suppression (where the engineers who raise concerns are punished, sidelined, or ignored, removing the organisation's primary corrective mechanism); regulator capture (where the regulatory body that should impose external discipline becomes dependent on the regulated industry for expertise, funding, or political support).

The discipline that studies these patterns is partly sociology (Vaughan's \emph{The Challenger Launch Decision} \cite{hist:vaughan1996} is the foundational ethnographic work), partly organisational science (Snook's \emph{Friendly Fire} on the Black Hawk shootdown; the high-reliability organisation literature of Weick, Sutcliffe, La Porte, Roberts), partly accident-investigation practice (the Rogers Commission on Challenger \cite{acc:nasa-challenger}, the CAIB on Columbia \cite{acc:caib-columbia}, the House Committee on 737 MAX \cite{acc:house-737max-2020}, all of which dedicate substantial sections to organisational analysis), and partly the practitioner literature of safety-culture assessment (INPO in nuclear; the IAEA safety-culture frameworks; the various oil and gas industry initiatives after Macondo and Buncefield).

\begin{principle}[Organisations are part of the system]
A safety analysis that does not include the organisational decision processes that govern operations, maintenance, change management, and response to weak signals analyses an incomplete system. The engineering elements may be correctly analysed and the resulting safety case still be wrong if the organisation will not enact what the analysis specified.
\end{principle}

\section{Vaughan's normalisation of deviance}\pathtag{core}

Diane Vaughan's \emph{The Challenger Launch Decision} \cite{hist:vaughan1996} reconstructs the years of organisational decision-making that preceded the Challenger accident on 28 January 1986 \cite{acc:nasa-challenger}. The central conceptual contribution is the concept of normalisation of deviance: a pattern by which a signal that initially registers as a deviation from expected behaviour is reclassified, through repeated observation without consequence, as within the acceptable operating envelope.

The Challenger O-ring sequence is the canonical example. The solid rocket booster joints had been observed, on multiple flights before Challenger, to experience "blow-by" of hot combustion gases past the primary O-ring seal during the early seconds of ignition. The blow-by was not the designed behaviour; the joints were specified to seal hermetically. On the first observations, engineers at Morton Thiokol (the SRB manufacturer) and at NASA's Marshall Space Flight Center treated the observations as anomalous. The flights had occurred without catastrophic failure; the blow-by had been arrested by the secondary O-ring; the joint had ultimately sealed before the booster's case pressure could escape. Repeated observations, each ending in successful flight, gradually shifted the organisational frame: the blow-by was no longer "anomalous" but "expected behaviour within the joint's actual (rather than specified) operating envelope." The accident occurred on the flight whose ambient temperature drove the primary O-ring's resilience past the point at which it could seal in time, and the secondary O-ring (compromised by the blow-by) could not arrest the leak.

Vaughan's account is not the simple "obvious warning ignored" framing that the Rogers Commission's executive summary supported. Vaughan's reconstruction shows that the engineers and managers involved made decisions that were locally rational within the organisation's evolved interpretive frame: each individual flight's data point was interpreted within a frame already shifted by the previous flights; the shift was incremental and was supported at each step by the engineering analyses (re-derived margins, re-justified flight-readiness reviews) that the organisation produced. The accident is the consequence of the cumulative shift, not of any single decision to ignore a known warning.

The mechanism generalises beyond Challenger. The 737 MAX MCAS development \cite{acc:house-737max-2020} exhibits the same pattern: the system's authority was expanded incrementally; each expansion was approved within an organisational frame that the previous expansion had shifted; the single-AoA architecture, adequate for the original limited authority, became inadequate for the expanded authority without the safety-assessment review that would have caught the inadequacy being triggered. The Therac-25 \cite{paper:leveson-turner1993} exhibits a similar pattern in the maintenance and modification of the software: known anomalies were closed without root-cause analysis because each anomaly could be explained away within the operating frame.

The defensive responses are several. Calibrated re-derivation: when an anomaly is observed, the analysis that established the operating envelope is re-derived from the underlying physics (not from the previous flight's success), and the deviation is evaluated against the re-derived envelope. External review: the organisation cannot be the sole judge of whether its frame has shifted; periodic independent review, with the explicit charter to question the operating frame, is a structural check. Recording of dissent: engineering objections that are overruled are recorded with their rationale, both for accountability and to surface the frame's drift over time. Calibration against absolutes: certain limits (temperature operating envelopes, structural margins, sensor accuracy requirements) are treated as absolute, with deviation requiring formal change control rather than operational accommodation.

\begin{archetype}[failure]
Normalisation of deviance is the archetype's slow-time expression. A signal arrives that should warn; the warning fails not by the signal's absence but by the organisation's interpretive shift over time. The accident is the moment the shifted interpretation meets the unforgiving physics that did not shift.
\end{archetype}

\section{Production pressure and safety drift}\pathtag{core}

Production pressure is the engineering term for the schedule, cost, and output incentives that organisations apply to operational activity. In small amounts, production pressure is the discipline that prevents endless cost growth and infinite schedule slip; in excess, it erodes the safety margins that engineering analysis specified. Safety drift is the consequence: the actual operating envelope shifts from the specified envelope, the changes are small and incremental, and the organisation does not notice the drift until an accident occurs that the original envelope would have prevented.

The mechanisms by which production pressure produces drift:

\textbf{Deferred maintenance.} A scheduled maintenance action is postponed because the production schedule does not afford the downtime. The postponement is recorded as a planned future action; the planned future does not arrive. The equipment operates beyond its specified maintenance interval; failures that the maintenance would have prevented become more probable. Aging equipment compounds this; the deferral that was acceptable for new equipment is consequential for old equipment.

\textbf{Procedural shortcuts.} A procedure step is skipped because experience has shown it is unnecessary in normal operation. The shortcut is initially individual (one operator's habit); it spreads informally; it becomes the de facto procedure. The shortcut works in most cases (the procedure step was conservative); it fails in the cases the procedure step was specifically designed to catch.

\textbf{Tolerance relaxation.} A measurement is observed to fall outside specification; rather than rework or rejection, the measurement is reviewed and accepted under a "minor non-conformance" classification. The classification is initially exceptional; repeated use makes it routine; the de facto specification widens. The Bhopal MIC tank temperature was operated above its specified maximum for extended periods preceding the accident \cite{acc:icmr-bhopal-2004}; the refrigeration system that would have held it at the specified temperature had been disabled to save operating cost. The de facto operating envelope had widened from the design specification.

\textbf{Capability erosion.} Critical operational capabilities (skilled maintenance technicians, experienced operators, redundant equipment, trained emergency-response teams) are reduced for cost savings. The reduction is small at each step; the cumulative effect is a brittleness that becomes visible only when the system encounters conditions the reduced capability cannot handle. The pre-accident Bhopal plant's safety systems had been progressively reduced, the trained workforce diminished, the spare parts inventory cut, in response to production-cost pressures \cite{acc:icmr-bhopal-2004}.

\textbf{Time compression.} Activities that historically required a particular duration are compressed (the launch window must be met; the production schedule must be hit; the maintenance turnaround must be shortened). The compression squeezes out the contingency time that catches errors and the review time that catches anomalies. The Boeing 737 MAX development was conducted under schedule pressure to match the Airbus A320neo's market introduction; the House Committee report identifies time compression as a contributing factor to the gaps in MCAS's safety analysis \cite{acc:house-737max-2020}.

\textbf{Selective reporting.} When metrics drive behaviour, the metrics get reported; when production targets and safety-incident counts both exist, the production targets get hit and the safety-incident counts get suppressed (incidents are reclassified, near-misses are not recorded, root-cause analyses are foreshortened). The metric distortion makes the organisation's safety state look better than it is; the safety drift continues unobserved.

Safety drift is difficult to detect because each individual step is small and each is locally justified. The defensive responses are at the organisational level: tracking of operating-envelope drift over time as a metric in its own right; periodic re-derivation of safety margins as if from scratch; protected funding for capabilities that are not directly productive; independent auditing that compares actual operations against specifications; structural separation of production and safety reporting lines.

\section{Whistleblower dynamics}\pathtag{standard}

The engineers who recognise that the organisation is drifting toward an accident, who raise concerns, who refuse to sign off on inadequate analyses, are the organisation's primary corrective mechanism. Their treatment by the organisation determines whether the mechanism functions. In the canonical accident cases, the mechanism failed.

The Challenger launch decision was made over the explicit objections of Morton Thiokol engineers who had argued, on the evening before launch, that the low ambient temperature would compromise the O-ring seal \cite{acc:nasa-challenger,hist:vaughan1996}. The teleconference between Thiokol and Marshall Space Flight Center management ended with Thiokol management reversing its initial recommendation against launch under pressure from Marshall; the engineers' objections were recorded but not acted on. Roger Boisjoly's post-accident testimony before the Rogers Commission documented the dynamics; his subsequent career inside Thiokol was difficult; he became a recurring speaker on engineering ethics for the remainder of his life.

The 737 MAX certification was conducted against the explicit concerns of Boeing engineers and FAA authorised representatives who, the House Committee report documents, raised concerns about the MCAS authority, the single-AoA architecture, and the absence of pilot documentation \cite{acc:house-737max-2020}. The concerns were addressed inadequately or escalated insufficiently; the organisational decision-making process did not provide a mechanism for the concerns to halt the program.

The Volkswagen diesel defeat device was implemented by engineers who knew the device was a regulatory evasion; the post-accident accountability spread up the organisation but stopped well short of the senior executives who had set the production targets the defeat device was implemented to meet \cite{acc:epa-vw-nov-2015}.

The patterns:

\textbf{Concern raised; organisational frame interprets as resistance.} Engineers who raise concerns in good faith are interpreted by managers as obstructionist, as failing to be team players, as not understanding the larger picture. The frame distorts the signal: a substantive engineering concern is heard as a personnel-management problem.

\textbf{Concern escalated; the escalation chain absorbs it.} The concern reaches a manager; the manager addresses it (perhaps by re-tasking analysis, perhaps by reassurance, perhaps by minor procedural change) without escalating further; the concern is not transmitted to the level of decision that the concern's substance warrants. The escalation chain acts as a filter that progressively reduces the signal's amplitude.

\textbf{Concern escalated to the top; the top decides against.} The concern reaches the responsible decision-maker; the decision-maker, weighing the concern against the production pressure they are themselves under, decides to proceed. The decision may be recorded with a documented justification; the justification is later read against the accident outcome and found inadequate. The Challenger launch decision had this character.

\textbf{Concern raised externally (whistleblower).} An engineer, having exhausted internal channels, raises the concern externally (to regulators, to media, to professional bodies). The external escalation is a last-resort step that carries professional and personal cost; whistleblower-protection laws have evolved to reduce the cost but do not eliminate it. The organisation's response to an external whistleblower tends to focus on the breach of internal channels (the whistleblower violated confidentiality, undermined the team) rather than on the substantive concern; the substantive concern is dealt with only when the external attention compels it.

The defensive responses are structural: protected channels for raising safety concerns (independent ombuds offices; safety-reporting systems with assurance of non-retaliation; explicit policies that prohibit retaliation); explicit escalation requirements (concerns of certain classes must be escalated to specified levels; the receiving level must record the resolution); independent technical review (engineering concerns can be reviewed by technical authorities outside the program management chain); regulatory channels (regulators provide a channel that bypasses the organisation entirely). None of these are sufficient alone; their combination, with active organisational support for their use, is the institutional protection of the corrective mechanism that whistleblowers represent.

\section{The role of regulator capture}\pathtag{standard}

Regulator capture is the dynamic by which a regulatory body, charged with imposing external discipline on a regulated industry, becomes dependent on the industry for expertise, funding, political support, or career pipelines, and in consequence its discipline is reduced. The dynamic was named by economist George Stigler in 1971 in the context of economic regulation; the phenomenon applies in technical regulation as well, with substantial consequences for accident probability.

The mechanisms by which technical regulator capture occurs:

\textbf{Expertise dependence.} The regulated industry employs the technical experts; the regulator does not employ enough of them. The regulator must rely on industry-supplied analyses, industry-developed standards, industry-employed authorised representatives to perform certification work. The FAA's Organisation Designation Authorisation (ODA) arrangement with Boeing is the canonical recent case in aerospace; the House Committee identified the arrangement as a contributing factor in the 737 MAX certification gaps \cite{acc:house-737max-2020}. The arrangement's intent was efficient allocation of scarce regulatory engineering capacity; the consequence was that the technical depth of the certification of MCAS rested on Boeing-employed engineers whose direct loyalty was to the program management chain.

\textbf{Funding dependence.} The regulator's budget depends on the regulated industry's fees or on political support that the industry can affect. A regulator that imposes costly requirements on the industry faces budget consequences; the institutional incentive moderates the discipline.

\textbf{Career pipeline.} Regulator staff move to industry positions during their careers; industry staff move into regulator positions. The movement is often beneficial (it brings practical experience into the regulator and regulatory understanding into the industry), but it also produces the "revolving door" dynamic in which current regulators anticipate future industry employment and current industry staff have past regulatory relationships. Neither dynamic is corruption; both shift the centre of gravity of regulatory behaviour toward accommodation.

\textbf{Industry-led standard-setting.} The technical standards against which the regulator measures compliance are developed by industry-led standards bodies (ASME, IEEE, SAE, ISO, ASTM). The standards reflect the industry's collective engineering judgment; they also reflect the industry's collective interest in achievable requirements. A standard developed by an industry consortium tends not to be substantially more stringent than the consortium members can comfortably meet.

\textbf{Political pressure.} The regulated industry is politically organised; political pressure on the regulator (through legislative oversight, through funding decisions, through executive-branch direction) constrains the regulator's discipline. The pressure is rarely explicit but its effect is real.

The institutional responses include: protected regulator funding (independent appropriation; user-fee structures that decouple budget from political pressure); independent technical authority (NTSB-style independence; the NRC's relative independence from the executive branch); explicit conflict-of-interest disclosure for regulator staff and standards-body participants; independent post-event reviews of regulator performance (the House Committee report on 737 MAX is structurally an instance of this); and the structural separation of certification (which establishes initial compliance) from continued surveillance (which monitors ongoing compliance), so that the certification's depth is not the sole defence against drift.

\section{Case in full depth: Challenger and Columbia, Boeing 737 MAX, Bhopal, Chernobyl}\pathtag{core}

\subsection*{Challenger and Columbia}

The Challenger accident on 28 January 1986 killed seven astronauts when the right Solid Rocket Booster's aft field joint failed during ascent, releasing combustion gases that breached the External Tank and caused the vehicle's aerodynamic break-up at $T+73$ seconds \cite{acc:nasa-challenger}. The Rogers Commission identified the technical mechanism (O-ring loss of resilience at low temperature) and the organisational mechanism (decision-making process that proceeded with the launch over the explicit objections of the engineers who had analysed the temperature effect). Vaughan's subsequent reconstruction \cite{hist:vaughan1996} identified the longer-time-scale normalisation-of-deviance pattern by which the joint's incremental performance shortfall had been progressively accepted as within the operating envelope.

The Columbia accident on 1 February 2003 killed seven astronauts when the orbiter disintegrated on re-entry due to wing-leading-edge damage that had been sustained during ascent (a piece of External Tank foam insulation struck the leading edge at approximately $T+82$ seconds, creating a hole through which super-heated re-entry plasma later entered) \cite{acc:caib-columbia}. The CAIB's analysis is notable for its explicit treatment of the organisational mechanism as a co-equal cause of the accident, not a contributing factor. The CAIB found that foam-debris strikes had occurred on multiple prior flights, had been increasingly tolerated as normal, had been classified as a "turnaround issue" (maintenance burden) rather than a flight-safety issue, and that during the Columbia mission's flight phase the engineers who recognised that the foam strike might have caused critical damage were unable to obtain on-orbit imagery (which would have confirmed the damage and might have enabled rescue or repair options) because the program management did not authorise the imagery request.

The CAIB's central conclusion was that the organisational causes of Columbia were structurally identical to those of Challenger: production pressure (the schedule for the Space Station's completion), normalisation of deviance (foam strikes treated as routine), suppression of engineering concern (the request for imagery was deflected), and a safety-management structure that lacked independence from program management. The conclusion is one of the most direct organisational-failure-recurrence statements in any major accident report; the recommendation was the establishment of an independent NASA safety and engineering technical authority with budget and authority separate from the program management chain.

\subsection*{Boeing 737 MAX}

The 737 MAX accidents (Lion Air 610, 29 October 2018; Ethiopian 302, 10 March 2019) killed 346 people through the Maneuvering Characteristics Augmentation System (MCAS) failure analysed in Chapter 8 \cite{acc:house-737max-2020}. The House Committee report identifies the organisational mechanism in unsparing detail: a commercial premise that the MAX would require no simulator training for 737NG-rated pilots, which shaped what could be documented to operators; schedule pressure to match the Airbus A320neo's market introduction, which compressed development time; an organisational culture that prioritised program execution over engineering concerns, which suppressed escalation of safety issues; an FAA delegation arrangement (ODA) that placed certification work in the hands of Boeing-employed authorised representatives whose loyalties were divided; and a safety-management process that did not require re-derivation of MCAS's safety analysis when its authority was expanded during development.

The 737 MAX is the canonical recent case for organisational-failure analysis in part because the House Committee report is so direct about the mechanism. The report identifies specific decisions, specific decision-makers, and specific cultural and structural patterns; it does not stop at "Boeing must improve its safety culture." The accountability outcomes (Boeing leadership changes; the \$2.5 billion criminal-fraud settlement; the ongoing FAA reform of ODA arrangements) are unusually direct consequences of an accident-investigation report; the underlying patterns are common to many of the other cases in this chapter.

\subsection*{Bhopal}

The Bhopal disaster of 2--3 December 1984 killed an estimated 3,800 immediately and tens of thousands over subsequent years through inhalation of methyl isocyanate (MIC) released from a Union Carbide India Limited pesticide plant in Bhopal, Madhya Pradesh \cite{acc:icmr-bhopal-2004}. The technical sequence began with water entering an MIC storage tank (likely through a misrouted maintenance operation; the exact mechanism remains contested), triggering a runaway exothermic reaction that overpressured the tank and released approximately 40 tons of MIC vapour through a stack into the populated area around the plant.

The organisational analysis identifies multiple chronic conditions of the pre-accident plant. The refrigeration system that would have held the MIC tanks at $0\,\degC$ (substantially slowing any contamination-induced reaction) had been disabled for over a year, reportedly to save operating cost. The flare tower that would have burned the released vapour was not operational. The vent gas scrubber that would have absorbed the released vapour was undersized for the actual release and had been out of service for maintenance. The plant's MIC inventory (60 tons in storage at the time of the accident) exceeded the inventory that conservative chemical-process safety practice would have allowed for the material's hazard class. The plant's workforce had been reduced; the trained operators who would have recognised the early signs of the runaway had been progressively replaced.

The corporate analysis (parent Union Carbide Corporation and Indian subsidiary Union Carbide India Limited) identified a pattern of progressive disinvestment in the plant: capital expenditure had been cut; the plant was not being modernised; safety equipment was not being maintained or replaced; the parent's safety standards (which were higher than the actual operating practice at Bhopal) were not enforced by the parent in its Indian subsidiary. The arrangement insulated the parent from direct liability for operational decisions while permitting the operational decisions that produced the accident. The post-accident legal and regulatory outcomes (the \$470 million settlement; the criminal proceedings against Union Carbide India personnel; the institutional changes in Indian process-safety regulation) addressed parts of the consequences but the underlying organisational pattern, of multinational disinvestment in operations in jurisdictions with weaker regulation, recurs in other cases (Texaco-Pemex relationships; the offshore drilling industry; the maritime shipping industry) and is not a problem any single accident's investigation has fully resolved.

\subsection*{Chernobyl}

The Chernobyl accident on 26 April 1986 killed 31 people in the immediate response and produced long-term consequences (cancer increases; exclusion zone) whose count is debated but is in the thousands. The technical sequence was a power excursion during a low-power test of Unit 4's RBMK reactor; the reactor's positive void coefficient (a design property by which loss of coolant accelerated rather than decelerated the reaction) combined with the test conditions to produce an exponential power rise; the reactor disassembled; the graphite moderator ignited; the release continued for ten days \cite{acc:iaea-insag1-1986,acc:iaea-insag7-1992}.

The IAEA's initial INSAG-1 report attributed the accident substantially to operator violations of operating procedures; the INSAG-7 update in 1992 substantially revised the assessment, identifying the reactor's design properties (the positive void coefficient, the control-rod displacer-tip design that caused brief positive reactivity insertion on rod insertion, the reactor's inadequate diversity of shutdown mechanisms) as primary causal factors and the operator actions as conditions that exposed the design flaws. The shift in the assessment is itself notable: the initial framing protected the design and the regulatory authority that approved it; the revised framing acknowledged that the design had inherent properties that any operator might have encountered.

The organisational analysis identifies multiple Soviet-system-specific features and a set of features common to other organisational-failure cases. The Soviet-specific features include: the absence of an effective safety-regulatory structure independent of the nuclear-power operating organisations; the secrecy that prevented the international nuclear-safety community from reviewing the RBMK design; the production targets and political pressures that drove test schedules and the conditions under which they were conducted; the absence of an effective whistleblower channel for the operators and engineers who had concerns about the test. The common features include: normalisation of deviance (the test had been postponed multiple times; the conditions under which it was conducted differed from the test plan in ways that had not been formally reviewed); production pressure (the test was scheduled to be completed before a planned maintenance shutdown to avoid further delay); and the suppression of engineering concern (the operators on shift had not been adequately briefed on the test plan; their concerns during the test were overruled by the test director).

The post-Chernobyl institutional response included: the establishment of the World Association of Nuclear Operators (WANO) as an industry peer-review body (modelled on INPO, which the Kemeny Commission had recommended after Three Mile Island); the IAEA Convention on Nuclear Safety (1994), which committed signatory states to specific regulatory and operational practices; the development of safety-culture frameworks as an explicit subject of nuclear-industry attention; and the eventual decommissioning of the RBMK reactors that lacked the post-Chernobyl modifications.

\begin{principle}[Engineering responsibility includes the organisation]
The engineer's responsibility for system safety extends to the organisational decisions that govern operation, maintenance, and change. The engineer who designs a safety system that the organisation will not maintain has designed a system that will not provide safety. The engineering judgment about whether to proceed with a design includes judgment about whether the organisation that will operate it will operate it as designed.
\end{principle}

\begin{project}[Five-page memo on an organisational failure]
Select one major organisational engineering failure (Challenger; Columbia; Bhopal; Chernobyl; Boeing 737 MAX; Deepwater Horizon; Volkswagen Dieselgate; Boeing 787 battery; Therac-25; or another with comparable depth in the literature). Read the primary accident investigation report and one substantive secondary analysis. Write a five-page memo, addressed to a hypothetical engineering manager in a similar industry, that answers: (1) What was the technical mechanism? (2) What was the organisational mechanism? (3) What did the engineering profession learn from the accident, in the sense of practices, standards, training, or organisational structures that changed because of it? (4) What did the engineering profession not learn, in the sense of patterns that recur in subsequent accidents that the lesson would have prevented? (5) What two specific changes do you recommend, in your reader's organisation, to reduce the probability of a similar failure?

\textbf{Hazard.} Engineering failures of this scale involve substantial loss of life. The memo's tone should be the engineering register, factual and constructive; it should not minimise the human cost. The point of the exercise is to identify what the engineering profession can do with the case, not to assign blame retrospectively.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
The Challenger O-ring sequence shows blow-by observed on approximately one-third of pre-Challenger flights \cite{acc:nasa-challenger}. Treating each flight as an independent Bernoulli trial with $p_{\text{blow-by}} = 0.33$ and $p_{\text{catastrophic given blow-by}} = q$, derive the expected number of flights before a catastrophic outcome as a function of $q$. For what value of $q$ does the expected number equal the actual number (approximately 24)?
\end{exercise}

\begin{exercise}
The 737 MAX MCAS authority was expanded during development; the original limited authority might have been $0.5\,\text{deg}$ per activation, and the expanded authority became $2.5\,\text{deg}$ per activation \cite{acc:house-737max-2020}. Compute the fractional change. Discuss what threshold should have triggered a re-derivation of the safety assessment.
\end{exercise}

\begin{exercise}
The Bhopal MIC inventory was approximately 60 tons at the time of the accident; the actual release was approximately 40 tons \cite{acc:icmr-bhopal-2004}. If best-practice chemical-process-safety inventory limits for MIC are on the order of 1 ton, compute the inventory-ratio overshoot. Discuss the implications for the magnitude of the release.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Model the normalisation-of-deviance dynamic as a random walk: each successful flight shifts the "acceptable" operating envelope by a small amount $\delta$ toward the actual observed behaviour; the actual envelope (set by physics) is fixed at $E$. Derive the expected number of flights before the perceived envelope reaches $E$ as a function of the initial gap and $\delta$.
\end{exercise}

\begin{exercise}
For the production-pressure / safety-drift dynamic, model the operating-margin reduction as a feedback loop in which (a) margin reductions reduce cost; (b) cost savings increase production; (c) production targets drive further margin reductions. Identify the conditions under which the loop is stable (margin remains bounded) versus unstable (margin drives to zero).
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the cost (lives, dollars, regulatory consequences) of the Challenger accident \cite{acc:nasa-challenger} and the cost of the safety-management changes the Rogers Commission recommended. Estimate the ratio.
\end{exercise}

\begin{exercise}
Estimate the fraction of major accident reports (NTSB, CSB, ATSB, BEA, AAIB) that identify production pressure, normalisation of deviance, or organisational decision failures as substantive causes. Discuss the implication for organisational-failure analysis as a core engineering discipline.
\end{exercise}

\begin{exercise}
Estimate the fraction of the 737 MAX MCAS development cost that re-running the system safety assessment after the authority expansion would have added \cite{acc:house-737max-2020}. Compare to the eventual cost of the accidents and the grounding.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design an organisational structure for a hypothetical aerospace manufacturer that provides explicit protection for engineers raising safety concerns. Specify the channels, the protections, the escalation requirements, and the audit mechanisms.
\end{exercise}

\begin{exercise}
Design a safety-management metric system that resists the selective-reporting failure mode (where production metrics are reported accurately and safety metrics are suppressed). Specify what is measured, how it is reported, who reviews it, and what consequences flow from the reports.
\end{exercise}

\begin{exercise}
Design a post-Columbia-style organisational reform for a hypothetical organisation in a different industry (a hospital network, a software platform company, a chemical-processing firm). Identify the structural changes that map to the CAIB recommendation of an independent safety and engineering technical authority \cite{acc:caib-columbia}.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: the engineer who signs off on a design under organisational pressure they believe will lead to an accident is professionally responsible for the accident, regardless of the organisational structure that produced the pressure. Argue the opposite position. Identify what the engineering profession's codes of ethics actually require.
\end{exercise}

\begin{exercise}
The CAIB found that the organisational causes of Columbia were structurally identical to those of Challenger \cite{acc:caib-columbia}. Discuss why the post-Challenger organisational reforms did not prevent Columbia. Identify what the post-Columbia reforms changed that the post-Challenger reforms did not.
\end{exercise}

\begin{exercise}
Discuss the regulator-capture analysis of the 737 MAX certification \cite{acc:house-737max-2020}. Identify the structural conditions that made capture possible. Evaluate whether the post-737-MAX reforms address those conditions or move them.
\end{exercise}

\begin{exercise}
Argue: the post-Bhopal multinational disinvestment pattern (parent-corporation insulation from subsidiary operations in lower-regulation jurisdictions) is itself a failure-producing organisational structure that the engineering profession should refuse to participate in. Identify the practical constraints on the position.
\end{exercise}
