\chapter{Maintenance, observability, life-cycle}
\label{vol09:ch18}

\epigraph{The maintainer's job is to make sure the next failure has not happened yet.}{}

\chapmeta{Half-life: durable. Archetypes: balance, failure, scaling. Exercise target: 25. Project track: physical.}

\noindent
Maintenance is the engineering activity that keeps systems operating across their useful life. The discipline addresses what fails, when it fails, how to anticipate it, how to repair it, when to replace components, when to retire systems entirely. For most engineering systems, total maintenance cost over the life cycle exceeds the original construction cost; for many, it exceeds by a factor of several. Despite this, maintenance is often the least-taught topic in engineering curricula and the least-glamorous topic in engineering practice. This chapter develops why maintenance matters, the standard maintenance strategies, condition monitoring, reliability-centred maintenance, observability as the foundation for maintenance, life-cycle costing revisited from the maintenance perspective, decommissioning and end-of-life management, and the characteristic failure of maintenance neglect as a root cause of accidents.

\section{Why maintenance is engineering's largest expense and least taught topic}\pathtag{standard}

The total life-cycle cost of an engineering asset is dominated by maintenance for most asset classes. A typical industrial pump: $\$10{,}000$ purchase price, $\$1{,}000/\text{year}$ maintenance, $20$-year life. Total maintenance cost: $\$20{,}000$, twice the purchase price. A typical commercial aircraft: $\$50\text{M}$ to $\$300\text{M}$ purchase price, $\$5\text{M}$ to $\$20\text{M}$/year maintenance, $25$-year life. Total maintenance cost: $\$125\text{M}$ to $\$500\text{M}$, often exceeding purchase price.

The same pattern applies across infrastructure (a $\$100\text{M}$ bridge has $\$2\text{M}$/year maintenance), buildings ($\$10\text{M}$ building, $\$200\text{k}$/year maintenance), and software (the famous claim that software maintenance is $70\%$ of total cost). The maintenance burden is not a marginal afterthought; it is the dominant economic reality of operating engineered systems.

Despite the economic importance, maintenance is poorly taught in engineering curricula. The reasons: maintenance is not as intellectually elegant as design; the topics span many specific disciplines; the textbook examples are less compelling than design examples; the academic prestige is lower. The result: graduating engineers know how to design but not how to maintain; the maintenance burden is borne by practitioners who learn on the job.

This chapter cannot fix the curriculum gap, but it can present maintenance as a serious engineering discipline. The discipline has its own concepts (reliability-centred maintenance, condition monitoring, predictive maintenance), its own analytical tools (failure-rate analysis, life-cycle costing, asset-management theory), and its own engineering practice. The discipline rewards systematic attention; engineering organisations that take maintenance seriously have measurably better operational outcomes than those that do not.

\begin{archetype}[failure: maintenance as ongoing failure prevention]
Maintenance is the ongoing engineering response to the universal failure tendency of all systems. Components wear; chemistry corrodes; materials fatigue; software bugs accumulate; configurations drift; people forget. Maintenance is the disciplined intervention that pushes back against this tendency. The art is doing enough to keep failure rates acceptable without spending more than the asset's value justifies. The engineering judgement is the right balance between maintenance investment and failure risk.
\end{archetype}

\section{Maintenance strategies: corrective, preventive, predictive, prescriptive}\pathtag{core}

The standard maintenance strategies, in increasing order of sophistication:

\textbf{Corrective (reactive) maintenance:} fix what breaks. The cheapest strategy per maintenance action but the highest in total cost (unscheduled downtime, secondary damage, safety risk). Suitable for non-critical components with low failure consequence and easy replacement.

\textbf{Preventive (scheduled) maintenance:} replace or service on a schedule, before failure. The schedule is based on manufacturer recommendations, industry standards, or historical data. Reduces unscheduled failures at the cost of some unnecessary maintenance. Suitable for components with known wear patterns and where unscheduled failure is costly.

\textbf{Predictive (condition-based) maintenance:} measure the component's condition; intervene when condition degrades but before failure. Requires condition-monitoring sensors and analysis. More efficient than preventive (no unnecessary maintenance) at the cost of monitoring investment. Suitable for high-value components with measurable degradation patterns.

\textbf{Prescriptive maintenance:} predictive maintenance combined with optimisation of the maintenance plan. The system not only predicts failure but recommends when, what, and how to maintain optimally. Modern data-driven approaches (machine learning on operational data) increasingly support prescriptive maintenance. Suitable for high-value, complex systems with rich operational data.

The choice of strategy depends on the component: criticality (consequence of failure), failure pattern (random vs wear-out), measurability (can condition be monitored), cost (of monitoring, of unscheduled failure, of preventive replacement). A single asset may use different strategies for different components: corrective for low-value parts, preventive for moderate-value parts, predictive for high-value or safety-critical parts.

The maintenance plan documents the strategy for each component: what is done, when, how, by whom, with what records. The plan is the engineering output of the maintenance discipline; it is the basis for resource planning, training, and verification.

\section{Condition monitoring}\pathtag{standard}

Condition monitoring measures the state of components to detect degradation before failure. Standard methods:

\textbf{Vibration analysis:} rotating machinery exhibits characteristic vibration signatures; changes in the signature indicate developing faults (bearing wear, misalignment, imbalance, looseness, gear damage). Accelerometers and spectral analysis identify the issue and (often) the specific bearing or component. Vibration is the dominant condition-monitoring method for rotating machinery.

\textbf{Oil analysis:} lubricating oil collects evidence of wear (metal particles), contamination (dirt, water, fuel), and oil degradation (oxidation, viscosity change). Spectroscopic and microscopic analysis quantify each. Used in engines, gearboxes, hydraulics, transformers.

\textbf{Thermal imaging:} hot spots indicate friction, electrical resistance, or compromised insulation. Infrared cameras inspect electrical panels, motors, bearings, refractory. Inexpensive thermal cameras have made this method routine for electrical inspection.

\textbf{Ultrasonic emission:} cracks, leaks, and electrical arcing produce ultrasound. Sensitive ultrasound detectors find these in motors, valves, electrical equipment, structures. Acoustic emission monitoring is a research-grade extension for structural-health monitoring.

\textbf{Performance monitoring:} the component's performance (output power, efficiency, response time) reflects its condition. Degraded performance indicates a problem; identifying the cause may require additional diagnostics. Routine for power plants, process equipment, transportation systems.

\textbf{Visual inspection:} the original condition-monitoring method, still important. Standard procedures with checklists, photographs, and trending. Augmented by drone inspection for tall or inaccessible structures.

\textbf{Online sensors:} permanently installed sensors providing continuous monitoring. Vibration, temperature, current, pressure, flow. Connected to monitoring systems that alarm on anomalies and trend conditions over time.

The discipline of condition monitoring includes: sensor placement (where is condition most informative); sensor selection (which sensor for which condition); analysis (interpreting the sensor data); thresholds (what condition indicates intervention); response (what to do when intervention is indicated). Modern IoT and ML extend each step; the discipline is the same.

\section{Reliability-centred maintenance}\pathtag{standard}

Reliability-centred maintenance (RCM) is a systematic approach to selecting maintenance strategies. The method emerged from the airline industry in the 1960s (MSG-1, MSG-2, MSG-3 documents) and is now standard practice in many high-reliability industries.

RCM proceeds in seven steps for each component:

1. \textbf{What is the function of the component?} (e.g., the pump must deliver $50\,\text{L/min}$ at $3\,\text{bar}$).

2. \textbf{What can fail (functional failures)?} (e.g., delivers less flow, more pressure, no flow, fails to start).

3. \textbf{What causes each failure (failure modes)?} (e.g., impeller wear, motor bearing failure, blocked inlet).

4. \textbf{What are the effects of each failure?} (e.g., process disruption, safety implications, environmental impact).

5. \textbf{How critical is each failure?} (criticality categories: catastrophic, high, medium, low).

6. \textbf{Which maintenance strategy is appropriate for each failure mode?} (corrective for low-criticality random failures, preventive for wear-out modes, predictive for measurable degradation).

7. \textbf{What is the specific maintenance task?} (e.g., monthly visual inspection of impeller, annual bearing replacement, online vibration monitoring with alarm thresholds).

The RCM analysis is documented and reviewed periodically. The output is the maintenance plan that drives operational activity. Modern RCM tools integrate with asset-management systems for execution and tracking.

RCM's discipline includes addressing failures the manufacturer's recommendations may not cover: site-specific failure modes, operational changes that alter failure patterns, modifications that affect maintenance. The RCM analysis is the engineering output that the maintenance plan implements.

\section{Observability: what to measure to manage}\pathtag{standard}

Observability is the system property that internal states can be inferred from external measurements. The control-engineering concept (Chapter 3) generalises to the engineering of complex systems: what must be measured to understand what is happening.

For maintenance, observability addresses several questions:

\textbf{Component condition:} sensors that reveal the component's state (vibration, temperature, current, pressure).

\textbf{System performance:} measurements that show whether the system is meeting its objectives (throughput, efficiency, response time, error rate).

\textbf{Operating envelope:} measurements that show how the system is being used (load patterns, environmental conditions, fault counts).

\textbf{Maintenance impact:} records that show how maintenance affected the system (post-maintenance performance, failure rates after intervention).

The "what to measure" question is itself an engineering question. Too few measurements leave maintenance blind; too many produce data overload that the maintenance team cannot use. The standard discipline: identify the failure modes (from RCM), identify the indicators of each failure mode, instrument those indicators, define thresholds and alarm logic, train the team to act on the data.

Modern observability in software systems (cloud-native applications, microservices) has its own discipline (metrics, logs, traces) with significant overlap with industrial observability. The same principles apply: measure what matters, alarm on what is actionable, retain history for diagnostics.

\section{Life-cycle costing revisited}\pathtag{standard}

The life-cycle cost (Chapter 15) is revisited here from the maintenance perspective. The total cost over the asset's life is:

$\text{LCC} = C_{acquisition} + C_{operation} + C_{maintenance} + C_{renewal} + C_{disposal} - V_{salvage}$

The maintenance portion is often the dominant term. The standard analysis breaks maintenance into:

\textbf{Routine maintenance:} regularly scheduled activity (lubrication, cleaning, inspection). Predictable cost.

\textbf{Preventive replacement:} scheduled replacement of components before failure. Cost depends on the schedule.

\textbf{Corrective maintenance:} response to failures. Cost depends on the failure rate and the average cost per intervention.

\textbf{Predictive intervention:} condition-based replacement before failure but after detection. Cost depends on the monitoring investment and the resulting intervention rate.

\textbf{Major overhauls:} periodic refurbishment to extend useful life. Major capital expense at intervals.

\textbf{Maintenance support:} parts, tools, training, documentation, software. Often underestimated.

The discipline of life-cycle costing includes attention to the maintenance side: realistic failure rates, realistic intervention costs, realistic schedules. The design-time decisions affect maintenance cost; the engineer should make them with maintenance in mind.

Design for maintenance (DFM) and design for testability (DFT) are engineering disciplines that consider maintenance from the beginning: components are accessible, modular, replaceable, testable. The cost of DFM/DFT during design is small; the savings during operation are large. The discipline is most established in aerospace and military systems where the operational cost is dominant.

\begin{estimation}
\textbf{Question.} Estimate the total maintenance cost over the life of a typical industrial pump.

\textbf{Working.} A $5\,\text{kW}$ industrial centrifugal pump costs $\$5{,}000$ to purchase. Annual energy use: $5 \times 8000 \times 0.7 \approx 28{,}000\,\text{kWh/year}$ at $\$0.12/\text{kWh}$ = $\$3{,}360/\text{year}$ (typical operating profile). Maintenance: routine inspection $\$200/\text{year}$, annual bearing replacement $\$300/\text{year}$, seal replacement every $3$ years at $\$500$, impeller replacement every $5$ years at $\$1000$, occasional unscheduled failures $\$500/\text{year}$ on average, parts and supplies $\$200/\text{year}$. Total annual maintenance: $\$200 + \$300 + \$167 + \$200 + \$500 + \$200 = \$1{,}567/\text{year}$. Over $20$-year life: $\$31{,}340$ in maintenance, compared to $\$5{,}000$ purchase; ratio is $6\times$. Energy is the largest single cost; maintenance is $33\%$ of total operating cost. The engineering case for energy-efficient equipment and predictive maintenance is strong.
\end{estimation}

\section{Decommissioning and end-of-life}\pathtag{standard}

The end of an asset's life involves: decommissioning (taking it out of service), dismantling (physical removal), disposal (handling the materials), and remediation (restoring the site).

Engineering considerations:

\textbf{Decommissioning planning:} the plan is best made at design time, not at end of life. The structure's disassembly, the hazardous materials handling, the regulatory requirements should be considered from the beginning. Nuclear plants are extreme examples: decommissioning takes decades and costs hundreds of millions to billions.

\textbf{Materials recovery:} components and materials may be valuable. Recycling reduces disposal cost and environmental impact. Standard discipline for major equipment (recovered through industrial recyclers); ad hoc for smaller equipment.

\textbf{Hazardous materials:} asbestos, PCBs, mercury, heavy metals, radioactive materials must be handled per regulation. The cost can be substantial.

\textbf{Site remediation:} the soil, groundwater, structures left after dismantling may need remediation. The cost can equal or exceed the operating-life cost of the asset.

\textbf{Regulatory:} permits, inspections, certifications, notifications. The regulatory burden is significant for industrial, nuclear, and infrastructure assets.

The discipline of design for disposal (related to circular-economy principles) considers end-of-life from the beginning. Material choices, modular design, labelled materials, and accessible disassembly all support efficient decommissioning. The discipline is emerging in product design (electronics, automotive) and is mature in some infrastructure contexts.

\section{Failure: maintenance neglect as the root cause; the slow accident}\pathtag{core}

Maintenance failures have characteristic patterns.

\textbf{Deferred maintenance accumulating to catastrophic failure:} as discussed in the infrastructure chapter. The maintenance is delayed; the deterioration accelerates; the failure becomes inevitable. The 2018 Genoa Morandi bridge collapse, the 2005 Texas City refinery explosion (\cite{acc:csb-bp-texas-city-2007}), and many other accidents have deferred maintenance components.

\textbf{The slow accident:} the failure mode develops slowly over years; the warning signs are present but not acted upon; the eventual failure looks sudden but the trajectory was visible. The 1986 Challenger O-ring problem (\cite{acc:nasa-challenger}) had multiple flights of erosion before the catastrophic launch; the warnings were discussed but not acted on. The 1979 Three Mile Island reactor's stuck valve (\cite{acc:kemeny-tmi-1979}) had previously stuck open during incidents that were investigated but not fully resolved.

\textbf{Maintenance-induced failure:} the maintenance activity itself caused the failure. A bolt installed wrong, a seal left out, a calibration not redone, a software change not propagated. The Aloha Airlines Flight 243 fuselage failure (\cite{acc:ntsb-aar94-04}, 1988) had a maintenance component: the multi-decade cyclic load with inadequate inspection allowed fatigue cracks to develop and join. The Sayano-Shushenskaya hydroelectric station's catastrophic turbine failure (2009) involved maintenance and operational decisions that allowed degraded equipment to operate beyond safe limits.

\textbf{Inspection failure:} the inspection program existed but did not detect the developing failure. The inspector did not see what was there, or did not recognise its significance, or saw it and did not report it. The Hyatt Regency walkway collapse (\cite{acc:nbs-hyatt-1982}, 1981) involved design and construction errors but the inspection program did not catch them.

\textbf{Knowledge loss:} the engineers who designed and built the asset retire; institutional knowledge about its operation and maintenance is lost; new operators operate it without full understanding. The 1986 Chernobyl accident (\cite{acc:iaea-insag7-1992}) had elements of operating-knowledge loss: the test that triggered the accident involved operating modes that the design team had not anticipated.

\textbf{Spare parts unavailable:} the maintenance requires parts that are no longer manufactured; the equipment is operated past the point where it can be maintained; failure becomes inevitable. The discipline of long-term spare-parts management is critical for long-lived assets.

\textbf{Maintenance personnel inadequacy:} the maintenance team is too small, too inexperienced, or too overworked. The result is rushed work, missed inspections, and accumulating problems. Maintenance budget cuts often produce this pattern.

\textbf{Documentation drift:} the maintenance documentation does not match the as-built asset; the maintenance procedures do not address actual failure modes. The result is ineffective or counter-productive maintenance.

\textbf{Operator-maintenance handoff failure:} the operator does not communicate the asset's history to maintenance; maintenance does not document its work to the operator; the next operator is unaware of issues; failures recur or are made worse.

\begin{principle}[The maintenance-as-engineering principle]
Maintenance is engineering, not just operations. The discipline includes analysis (which failure modes, which strategies), design (the maintenance plan), execution (the work), measurement (the performance), and improvement (the closed loop). The engineering of maintenance is as serious as the engineering of the original design; in fact, more of the asset's life-cycle cost depends on it. The organisation that treats maintenance as a cost centre to be minimised, rather than an engineering discipline to be performed well, accepts the consequences in failures, accidents, and accelerated asset deterioration.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Design a maintenance schedule for a real piece of equipment]
\textbf{Objective.} Design a maintenance schedule for a real piece of equipment using reliability-centred maintenance principles. Justify each maintenance action with explicit failure-rate and consequence assumptions. Document the schedule for handover to a hypothetical operator.

\textbf{Method.} Choose a piece of equipment: a car you own, a household appliance, an HVAC system, an industrial machine you have access to, a software system. Identify the functions; enumerate failure modes; assess criticality and failure rates; choose maintenance strategies; specify maintenance tasks. Document the schedule: what is done, when, by whom, with what parts, with what records.

\textbf{Deliverables.} The equipment description; the function-failure-mode analysis; the maintenance strategy selection with rationale; the schedule with task descriptions; a written analysis of the cost-benefit trade-offs in the choices.

\textbf{Hazard.} The schedule is paper; implementation should follow the equipment's actual safety procedures.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A pump has failure rate $\lambda = 0.05/\text{year}$; preventive replacement is performed every $T$ years. Compute the optimal $T$ that minimises total expected cost, given preventive cost $\$500$ and failure cost $\$5{,}000$.
\end{exercise}

\begin{exercise}
A vibration alarm threshold is $4.5\,\text{mm/s}$ (ISO 10816-3, zone B/C boundary); the bearing's vibration trend is $0.3\,\text{mm/s/month}$. Compute the time to alarm.
\end{exercise}

\begin{exercise}
A maintenance team has $5$ technicians; each services $20$ pieces of equipment quarterly. Compute the team capacity and the resulting equipment portfolio served.
\end{exercise}

\begin{exercise}
An asset's annual maintenance cost is $\$10{,}000$, growing at $5\%/\text{year}$; replacement cost is $\$200{,}000$; discount rate is $5\%$. Compute the NPV of $20$ years of maintenance.
\end{exercise}

\begin{exercise}
A condition-monitoring sensor system costs $\$50{,}000$; predicted maintenance savings are $\$15{,}000/\text{year}$ for $10$ years; discount rate $5\%$. Compute the NPV.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the optimal preventive-maintenance interval as a function of failure rate, preventive cost, and failure cost.
\end{exercise}

\begin{exercise}
Derive the relationship between condition-monitoring sensitivity (early detection) and intervention timing for a component with measurable wear.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the total maintenance cost over the life of a commercial vehicle (truck, $1\text{M}\,\text{km}$ life).
\end{exercise}

\begin{exercise}
Estimate the staffing required for the maintenance of a small data centre ($100$ servers, $24/7$ operation).
\end{exercise}

\begin{exercise}
Estimate the cost-of-failure for a single failure event of a critical aerospace component without backup.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate the long-run cost of three maintenance strategies (corrective, preventive, predictive) for a component with Weibull lifetime; identify the strategy that minimises total cost.
\end{exercise}

\begin{exercise}
Simulate condition-based maintenance with stochastic deterioration and intervention thresholds; identify the optimal threshold.
\end{exercise}

\begin{exercise}
Simulate the long-run cost of operating an equipment fleet with various replacement-age policies.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design an RCM analysis for a small fleet of industrial pumps: functions, failure modes, criticalities, strategies, tasks.
\end{exercise}

\begin{exercise}
Design a condition-monitoring strategy for a wind-turbine gearbox: sensors, analysis, thresholds, response.
\end{exercise}

\begin{exercise}
Design the decommissioning plan for a small chemical-process facility: tasks, sequence, regulatory, cost.
\end{exercise}

\begin{exercise}
Design a maintenance-management database schema: assets, tasks, schedules, history, costs, performance.
\end{exercise}

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A team's maintenance budget has been cut $30\%$. State the engineering response.
\end{exercise}

\begin{exercise}
A team is choosing between preventive and predictive maintenance for a fleet of $1000$ pumps. State the engineering trade-offs.
\end{exercise}

\begin{exercise}
A team's maintenance team is overworked and missing scheduled tasks. State the engineering response.
\end{exercise}

\begin{exercise}
A team's condition-monitoring system has generated many alarms over six months; no actual failures have been averted. State the engineering response.
\end{exercise}

\begin{exercise}
A team's asset has just experienced an unexpected failure; the failure mode was not in the FMEA. State the engineering response.
\end{exercise}

\begin{exercise}
A team is asked to extend an asset's life beyond original design lifetime. State the engineering considerations: condition assessment, additional inspection, operating-envelope restriction, eventual replacement.
\end{exercise}

\begin{exercise}
A team's decommissioning of an old facility is over budget and behind schedule; hidden conditions have been encountered. State the engineering response.
\end{exercise}

\begin{exercise}
A team is asked to defend a maintenance recommendation that adds significant cost; the customer wants justification. State the engineering response and the role of life-cycle-cost analysis.
\end{exercise}
