\chapter{MEMS and microsystems}
\label{vol08:ch12}

\epigraph{Scale the machine down by a thousand and the physics changes; design for the new physics.}{}

\chapmeta{Half-life: medium-life. Archetypes: scaling, interface, failure. Exercise target: 28. Project track: physical.}

\noindent
Microelectromechanical systems (MEMS) are sub-millimetre machines fabricated by silicon micromachining. The MEMS revolution made high-quality inertial measurement available at consumer cost: every smartphone since the iPhone 4 (2010) carries a three-axis accelerometer and gyroscope on chips smaller than a grain of rice. MEMS pressure sensors enable airbags and altitude-aware fitness trackers; MEMS microphones replace electret capsules in consumer electronics; MEMS optical mirrors enable digital projectors and LIDAR; MEMS resonators provide frequency references in consumer electronics. This chapter develops the engineer's working understanding of why scaling matters in micromachines, the fabrication processes that produce MEMS, the major MEMS sensor families (accelerometer, gyroscope, pressure, microphone), the MEMS actuator catalogue, RF MEMS and bulk-acoustic-wave filters, microfluidics and lab-on-chip systems, optical MEMS, and the failure mechanisms that distinguish MEMS from macroscopic machines. The reader who finishes this chapter can specify and integrate MEMS devices and assess MEMS-related failure modes.

\section{What MEMS are and why scaling matters}\pathtag{core}

A MEMS device is a mechanical structure with feature sizes from microns to millimetres, fabricated by adapting integrated-circuit processes (photolithography, thin-film deposition, etching) to work in three dimensions. The structures move (cantilevers bend, beams flex, masses shift, membranes deflect, mirrors tilt), and the motion is sensed or driven electrically (usually capacitively).

Scaling changes which physical effects dominate. As linear dimensions shrink by a factor $\alpha$, masses scale as $\alpha^3$, areas as $\alpha^2$, surface forces (capillarity, electrostatic, Van der Waals) scale as $\alpha$, and natural frequencies scale as $\alpha^{-1}$. At MEMS scales:

\textbf{Surface forces dominate over gravity.} A MEMS cantilever sits on whatever surface forces hold it; gravity is a perturbation. Stiction (adhesion of a released structure to the substrate during fabrication or in service) is a leading failure mode that has no macroscale analogue.

\textbf{Resonance frequencies are high.} A $100\,\mu\text{m}$-long silicon cantilever resonates at $\sim 100\,\text{kHz}$; the same beam at $10\,\text{mm}$ scale would resonate at $\sim 1\,\text{kHz}$. High frequency means fast response, high bandwidth, and immunity to vibration at lower frequencies.

\textbf{Quality factor depends on environment.} A MEMS structure in vacuum has $Q$ in the thousands to millions; in air, $Q$ drops by orders of magnitude due to gas damping. The package atmosphere is a design parameter.

\textbf{Thermal time constants are short.} A MEMS thermal device equilibrates in microseconds to milliseconds; macroscale thermals take seconds to minutes. This enables fast thermal sensors and actuators, and constrains thermal stability of sensitive devices.

\textbf{Electrostatic actuation becomes practical.} Voltages of a few to a few hundred volts produce useful forces at MEMS gaps of a few microns; the same voltages at macroscale gaps are negligible. Capacitive sensing at the same scales detects sub-femtofarad changes that macroscale capacitive sensing cannot resolve relative to parasitic capacitances.

The MEMS engineer designs to the scaling: choosing the dominant physics, exploiting it, and mitigating the parasitic effects.

\begin{archetype}[scaling: MEMS as scaling-changed regime]
A MEMS device is not simply a small version of a macroscale device. The relative magnitudes of gravity, surface forces, thermal effects, and inertial effects change with scale, putting MEMS into a different regime. Designs that work at the macroscale (gravity-driven actuation, fluid pumps relying on inertia, oscillators with bulk inertia) usually do not transfer to MEMS; the MEMS catalogue is its own design space with its own dominant physics.
\end{archetype}

\section{MEMS fabrication: bulk, surface, LIGA}\pathtag{standard}

Bulk micromachining etches structures into the silicon substrate using anisotropic wet etchants (KOH, TMAH) or deep reactive-ion etching (DRIE). Anisotropic wet etching follows crystal planes, producing well-defined pyramidal pits and trenches; DRIE produces vertical sidewalls with high aspect ratios. Bulk micromachining is used for pressure sensor diaphragms, large inertial proof masses, and through-wafer vias.

Surface micromachining deposits layers of polysilicon and sacrificial silicon dioxide, patterns them, and releases the structures by removing the sacrificial layer (typically HF etch). The result is suspended polysilicon structures with micron features. Surface micromachining is the dominant MEMS process: Analog Devices' inertial sensors, Texas Instruments' DLP mirrors, and many others use it. The process maturity allows commodity-cost production.

LIGA (Lithographie, Galvanoformung, Abformung) uses X-ray lithography to define tall, high-aspect-ratio structures in PMMA, then electroplates metal into the resist patterns. The resulting metal structures can be hundreds of microns tall with sub-micron sidewall accuracy. LIGA is expensive and limited to specialised applications (high-quality micro-gears, high-aspect-ratio inertial masses).

Wafer-bonding (anodic, eutectic, fusion) joins multiple wafers to create sealed cavities, hermetic packages, or stacked structures. Pressure sensors with sealed reference cavities, gyroscope chambers under vacuum, and packaged microfluidic devices all use wafer bonding.

Modern MEMS production is integrated: a typical inertial sensor has the MEMS structure on one die, a CMOS readout ASIC on a separate die, and the two die are co-packaged or stacked. Some processes (TSMC's MEMS-CMOS integration) integrate both on one wafer; the choice depends on volume, cost, and process compatibility.

The package matters as much as the chip. MEMS packaging must protect the moving structure from contamination, control the atmosphere (vacuum, dry nitrogen, hermetically sealed), and survive the mechanical and thermal stresses of board mounting. Failure of the package leads to MEMS failure; packaging-induced stress changes the device's behaviour.

\section{MEMS sensors: accelerometer, gyroscope, pressure, microphone}\pathtag{standard}

\textbf{MEMS accelerometers} measure acceleration by sensing the displacement of a proof mass relative to the substrate. The capacitive accelerometer has a movable comb structure between fixed comb electrodes; under acceleration, the proof mass shifts, the capacitances change differentially, and the readout circuit measures the difference. Typical specifications: range $\pm 2$ to $\pm 200\,\text{g}$, noise density $50$ to $200\,\mu\text{g}/\sqrt{\text{Hz}}$, bandwidth $1\,\text{Hz}$ to several $\text{kHz}$, cost from cents to dollars. Piezoresistive accelerometers (older, simpler readout) and piezoelectric accelerometers (high frequency, no DC) are alternative technologies.

\textbf{MEMS gyroscopes} measure angular velocity by sensing the Coriolis force on a vibrating proof mass. The proof mass is driven into oscillation along one axis; rotation about a perpendicular axis produces a Coriolis force perpendicular to both, which displaces the mass and is detected capacitively. The vibrating-structure (or "tuning fork") principle was demonstrated commercially in the 1990s and now dominates. Specifications: range $\pm 100$ to $\pm 2000\,^\circ/\text{s}$, bias stability $1$ to $100\,^\circ/\text{hr}$, noise density $0.01$ to $1\,^\circ/\sqrt{\text{hr}}$, cost from a few dollars to hundreds (navigation grade).

\textbf{MEMS pressure sensors} measure pressure by sensing the deflection of a thin diaphragm. Piezoresistive sensing uses doped resistors on the diaphragm; capacitive sensing measures the gap change between the diaphragm and a back-plate. The reference can be absolute (vacuum sealed cavity), gauge (atmospheric reference), or differential (two-port). MEMS pressure sensors range from $0.1\,\text{kPa}$ (precision altitude) to $100\,\text{MPa}$ (hydraulic), with accuracy of $0.1$ to $1\%$ at cost of a few cents to a few dollars.

\textbf{MEMS microphones} measure sound pressure with a small diaphragm capacitively coupled to a back-plate. The signal is amplified by an integrated CMOS ASIC. MEMS microphones have largely replaced electret condenser microphones in smartphones, hearing aids, and wearable devices: better consistency, smaller size, easier surface mounting, $-26\,\text{dBV/Pa}$ sensitivity, $65\,\text{dB}$ SNR at $\$0.50$.

The MEMS sensor market grew with the smartphone (IMU as a free side-effect of phone design), the automotive market (airbag sensors, electronic stability control, tyre pressure monitoring), and increasingly the wearable and medical-device market. Annual unit volumes are in the tens of billions.

\section{MEMS actuators: comb drives, piezo, electrothermal}\pathtag{mastery}

\textbf{Comb drive actuators} use the electrostatic force between interdigitated comb fingers. Applying a voltage produces a force pulling the moving comb into the fixed comb. The force is proportional to voltage squared and to the number of finger pairs; typical forces are $\mu\text{N}$ at $50\,\text{V}$ for hundreds of finger pairs. Comb drives are used for resonant gyroscopes, optical scanners, and tunable filters. The force is non-contact (no friction), making the mechanism nearly wear-free.

\textbf{Parallel-plate electrostatic actuators} use two parallel plates with a voltage between them. The force is proportional to voltage squared and inverse to gap squared. Pull-in instability at $\sim 1/3$ of the initial gap limits the stable operating range. Parallel-plate actuators are used for switches, ratchet mechanisms, and digital micromirror devices.

\textbf{Piezoelectric MEMS actuators} use thin-film piezoelectric materials (typically AlN, ScAlN, or PZT) to convert applied voltage to strain. Forces are moderate (mN), displacements small ($\mu\text{m}$), bandwidth high (MHz). Piezo MEMS dominate ultrasonic transducers for medical imaging and consumer fingerprint sensors.

\textbf{Electrothermal actuators} use ohmic heating of a beam with non-uniform cross-section to produce bending. A current heats the thin section more than the thick section, the asymmetric expansion bends the beam, and the displacement is the actuation. Electrothermal actuators are simple, produce large displacements ($10\,\mu\text{m}$ from a $100\,\mu\text{m}$ beam) at low voltage, but consume significant power and have low bandwidth (heating and cooling times).

\textbf{Shape memory MEMS} use thin-film SMA (NiTi) to produce large displacements with thermal actuation. The actuators are slow but powerful.

The choice of actuator depends on the force, displacement, bandwidth, voltage, and power budget. The comb drive is the default for moving-mass MEMS; the parallel-plate switch is the default for digital MEMS; the piezo is the default for high-bandwidth small-displacement.

\section{RF MEMS and BAW/FBAR filters}\pathtag{mastery}

RF MEMS use micromechanical resonators as filters, oscillators, and switches at radio frequencies. The dominant commercial form is the bulk-acoustic-wave (BAW) filter, especially the thin-film bulk acoustic resonator (FBAR) developed by Avago/Broadcom and the solidly-mounted resonator (SMR) developed by Infineon. BAW filters dominate the LTE and 5G smartphone duplexer market; a modern smartphone contains $30$ to $50$ BAW filters in its RF front-end.

The BAW resonator is a thin-film piezoelectric (typically AlN or ScAlN) sandwiched between metal electrodes, supported either by acoustic isolation from the substrate (FBAR) or by a Bragg reflector (SMR). At the resonance frequency (set by the piezo thickness and acoustic velocity, typically $1$ to $5\,\text{GHz}$), the structure has very high Q ($1000$ to $4000$), making it the right element for narrow-band filters at sharp spectrum edges.

MEMS oscillators (SiTime, Discera) replaced quartz crystals in some applications, particularly small-form-factor and shock-tolerant uses. The MEMS resonator is integrated with a compensation circuit in a small package, giving stability of $10$ to $100\,\text{ppm}$ over operating range, with shock tolerance an order of magnitude better than quartz.

MEMS switches (still limited in commercial deployment, though research-mature) offer near-zero loss when on and very high isolation when off; their reliability under millions of cycles has limited adoption in high-volume products. The 5G era's mmWave electronics may renew the case for MEMS switches.

\section{Microfluidics: lab-on-chip, droplet microfluidics, organ-on-chip}\pathtag{mastery}

Microfluidic devices manipulate small volumes of fluid (nanolitres to microlitres) through channels with sub-millimetre dimensions. At these scales, viscous forces dominate over inertia (low Reynolds number), surface tension dominates over gravity, and diffusion is fast over short distances. Microfluidic flow is laminar; mixing requires explicit design.

Lab-on-chip devices integrate sample preparation, reaction, separation, and detection on a single chip. The PCR thermocycler and gel electrophoresis box of a molecular biology lab become a chip with heaters, channels, and integrated detection. Lab-on-chip platforms enable point-of-care diagnostics (low-cost HIV viral load test, malaria detection, blood typing) and high-throughput screening (drug discovery).

Droplet microfluidics generates and manipulates small droplets of one fluid suspended in another (water-in-oil typically). Each droplet is a sub-microlitre reactor. Single-cell sequencing, single-cell PCR, and digital droplet PCR exploit this: encapsulate one cell per droplet, react, count droplets. The single-cell genomics revolution depends on droplet microfluidics.

Organ-on-chip devices combine microfluidic channels with cultured cells to mimic the function of an organ. Lung-on-chip, gut-on-chip, kidney-on-chip, and liver-on-chip platforms enable pharmacological testing that historical cell-culture and animal-model alternatives cannot match. The field is research-active and beginning to inform regulatory submissions.

\section{Optical MEMS: DLP, MEMS mirrors, photonic integration}\pathtag{standard}

The digital light processor (DLP), Texas Instruments' MEMS mirror array, dominates projection displays. Each pixel is a micromirror that tilts $\pm 12^\circ$ between two states; modulating the time the mirror spends in each state produces grayscale. A DLP chip has $1$ to $8$ million mirrors at $5.4$ to $13.6\,\mu\text{m}$ pitch. The DLP architecture went from research curiosity (1987) to dominant projection technology by 2000.

MEMS scanning mirrors steer a laser beam in one or two axes. Applications include LIDAR for autonomous vehicles, retinal-display projection, optical-switch fabric for telecommunications, and barcode scanning. The scanning mirror is small, low-power, and shock-tolerant; the resonant mode (the mirror oscillates at its natural frequency, driven by small electrostatic or piezo force) gives high amplitude with low input. Quasi-static mirrors (driven open-loop to specific angles) suit LIDAR's pointing requirements.

Photonic integration combines optical waveguides, modulators, detectors, and lasers on silicon. The technology is mature for telecommunications transceivers (100 to 400 Gbps) and emerging for biosensing, LIDAR, and quantum-computing readout. Silicon photonics shares fabrication with CMOS, allowing co-integration with electronics.

\section{Failure: stiction, charge-trapping, packaging-induced stress}\pathtag{core}

MEMS failures have characteristic patterns distinct from macroscale electronics.

\textbf{Stiction:} a released MEMS structure sticks to the substrate or to an adjacent surface due to surface forces (capillarity during release, Van der Waals in operation). Stiction during release is mitigated by supercritical-CO$_2$ drying or vapour-phase HF release. Stiction in service is mitigated by surface coatings (hydrophobic monolayers, fluoropolymers), bump stops, and operating-condition control (electric-field-induced contact recovery). Stiction is the leading cause of MEMS yield loss in some processes.

\textbf{Charge trapping:} dielectric materials near the moving structure accumulate trapped charge under electrostatic operation, shifting the device's pull-in voltage, biasing the capacitive sense, and over time, causing drift in calibration. Charge trapping is exacerbated by humidity and high temperature. Mitigation: bipolar drive (alternating polarity to cancel trapping), shielded dielectrics, careful material selection. Charge trapping is the leading lifetime-limiting mechanism in many capacitive MEMS.

\textbf{Packaging-induced stress:} the package transmits temperature, humidity, and mechanical stress to the MEMS chip. The MEMS structure's behaviour shifts: zero-offset drifts, sensitivity changes, sometimes catastrophic failures. Mitigation: stress-isolating die-attach, careful PCB layout, package qualification testing. Packaging-induced stress is the leading cause of failure on field deployment when laboratory units worked fine.

\textbf{Particulate contamination:} a particle in the moving structure's path causes catastrophic failure. The wafer-clean and packaging cleanliness are part of process control; the package seal must keep particles out for life.

\textbf{Mechanical shock:} a sudden acceleration (drop, impact) can exceed the structure's elastic limit and crack or break the suspension. MEMS specifications include shock survival (typically $10{,}000\,\text{g}$ for $0.2\,\text{ms}$ for consumer sensors). Aerospace MEMS for higher shock environments use mechanical stops and over-range protection.

\textbf{Hermeticity loss:} packaged-vacuum MEMS (gyroscopes, oscillators) require vacuum for their Q; package leakage degrades Q, ultimately causing failure. The package qualification includes hermeticity testing (helium leak rate). Hermeticity degradation is a long-term reliability concern.

\section{Project}\pathtag{core}

\begin{project}[Characterise a smartphone MEMS accelerometer]
\textbf{Objective.} Characterise the MEMS accelerometer in a smartphone or a development-kit IMU. Measure: zero-g offset (and its temperature dependence), noise density, bandwidth, axis cross-talk, and temperature-dependent gain. Compare measured values to datasheet specifications.

\textbf{Method.} Use a smartphone with a data-logging app (PhyPhox, Sensor Logger, or similar) or a development board (Adafruit, SparkFun) with a logged-output IMU (LSM6DSV, ICM-42688, BMI270). Record acceleration with the device stationary at multiple orientations to extract zero-g offset and gravity calibration. Record while applying step changes (drop, tap) and oscillatory excitation (vibration table or audio speaker driving a holder). Use a temperature chamber if available, or a refrigerator/freezer/oven, to measure temperature dependence.

\textbf{Deliverables.} Measured zero-g offset per axis, noise spectrum from $0.1$ to $50\,\text{Hz}$, step-response bandwidth estimate, cross-axis sensitivity, temperature dependence of offset and gain (if measurable), and a written comparison to datasheet specifications.

\textbf{Hazard.} Drop testing risks damaging the phone or board; do drops onto soft surfaces only; keep the equipment secure during vibration testing.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A MEMS cantilever of length $200\,\mu\text{m}$, width $20\,\mu\text{m}$, thickness $2\,\mu\text{m}$, made of silicon ($E = 170\,\text{GPa}$, $\rho = 2330\,\text{kg/m}^3$), compute the natural frequency.
\end{exercise}

\begin{exercise}
A capacitive accelerometer has proof-mass $5 \times 10^{-9}\,\text{kg}$, spring constant $2.0\,\text{N/m}$. Compute the displacement at $1\,\text{g}$ and the natural frequency.
\end{exercise}

\begin{exercise}
A comb-drive actuator has $200$ finger pairs, gap $3\,\mu\text{m}$, finger thickness $10\,\mu\text{m}$, applied voltage $30\,\text{V}$. Compute the actuation force ($\varepsilon_0 \approx 8.85 \times 10^{-12}\,\text{F/m}$).
\end{exercise}

\begin{exercise}
A MEMS pressure sensor has a square diaphragm $200\,\mu\text{m}$ on a side, $5\,\mu\text{m}$ thick. Estimate the centre deflection at $10\,\text{kPa}$ pressure difference, using small-deflection plate theory.
\end{exercise}

\begin{exercise}
A MEMS gyroscope with $500\,\mu\text{m}$ proof mass drive amplitude, $20\,\text{kHz}$ drive frequency, $1\,\mu\text{kg}$ proof mass. Compute the Coriolis force at $100\,^\circ/\text{s}$ rotation.
\end{exercise}

\begin{exercise}
A parallel-plate capacitor has initial gap $5\,\mu\text{m}$, plate area $200 \times 200\,\mu\text{m}^2$. Compute the capacitance and the pull-in voltage if the suspension spring constant is $1\,\text{N/m}$.
\end{exercise}

\begin{exercise}
A MEMS microphone has sensitivity $-26\,\text{dBV/Pa}$ at $1\,\text{kHz}$. Compute the output voltage at $1\,\text{Pa}$ ($94\,\text{dB SPL}$).
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the natural frequency of a cantilever beam of length $L$, width $w$, thickness $t$, modulus $E$, density $\rho$, in the absence of damping.
\end{exercise}

\begin{exercise}
Derive the pull-in voltage and pull-in gap of a parallel-plate electrostatic actuator with linear suspension spring.
\end{exercise}

\begin{exercise}
Derive the Coriolis-force expression for a vibrating-structure gyroscope: $F_{coriolis} = 2 m \omega v$.
\end{exercise}

\begin{exercise}
Derive the squeeze-film damping coefficient of a parallel-plate structure in air, for small gaps.
\end{exercise}

\begin{exercise}
Derive the noise floor of a capacitive accelerometer, including Brownian motion noise of the proof mass and electronic noise of the readout.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the minimum detectable angular rate of a MEMS gyroscope with $1\,\text{mN/m}$ suspension stiffness, $10\,\mu\text{m}$ drive amplitude, $20\,\text{kHz}$ drive frequency, capacitive readout with $1\,\text{aF}$ resolution.
\end{exercise}

\begin{exercise}
Estimate the temperature coefficient of natural frequency for a silicon MEMS resonator (thermal expansion $\sim 2.6\,\text{ppm/K}$, modulus temperature coefficient $\sim -60\,\text{ppm/K}$).
\end{exercise}

\begin{exercise}
Estimate the BAW filter $Q$ required to achieve $1\%$ stopband attenuation $1\,\text{MHz}$ away from a $2.4\,\text{GHz}$ passband.
\end{exercise}

\begin{exercise}
Estimate the time for a $1\,\text{mm}$ aqueous droplet to mix by diffusion (diffusion coefficient $\sim 10^{-9}\,\text{m}^2/\text{s}$). Compare to a $10\,\mu\text{m}$ droplet.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate the modal frequencies of a MEMS accelerometer proof mass and identify the operating mode.
\end{exercise}

\begin{exercise}
Simulate the squeeze-film damping in a MEMS comb-drive structure, comparing air and vacuum operation.
\end{exercise}

\begin{exercise}
Simulate the temperature drift of a capacitive MEMS pressure sensor including thermal expansion and modulus change.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a capacitive accelerometer for $\pm 10\,\text{g}$ range, $10\,\mu\text{g}/\sqrt{\text{Hz}}$ noise, $500\,\text{Hz}$ bandwidth. State the proof mass, the spring stiffness, the gap, and the readout architecture.
\end{exercise}

\begin{exercise}
Design a MEMS pressure sensor for $0$ to $100\,\text{kPa}$, $0.25\%$ accuracy, capacitive readout. State the diaphragm dimensions, the back-plate geometry, and the temperature-compensation approach.
\end{exercise}

\begin{exercise}
Design a comb-drive resonant gyroscope architecture, including drive and sense modes, mode-matching strategy, and quadrature cancellation.
\end{exercise}

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A team's MEMS accelerometer shows $5\%$ unit-to-unit variation in sensitivity. State three engineering hypotheses and the corresponding investigations.
\end{exercise}

\begin{exercise}
A team's MEMS gyroscope shows bias instability $10\times$ higher than the datasheet. State three engineering causes and the corresponding actions.
\end{exercise}

\begin{exercise}
A team is choosing between a consumer-grade MEMS IMU ($\$2$, $5\,^\circ/\text{hr}$ bias) and an industrial MEMS IMU ($\$200$, $0.5\,^\circ/\text{hr}$ bias) for a $5$-minute autonomous navigation task. State the engineering trade-offs.
\end{exercise}

\begin{exercise}
A team's MEMS microphone has acceptable performance in initial tests and degrades after reflow soldering. State three engineering hypotheses.
\end{exercise}

\begin{exercise}
A team's MEMS pressure sensor drifts $100\,\text{Pa}$ in the first day of field deployment. State the engineering investigation, including the role of package outgassing and stress relaxation.
\end{exercise}

\begin{exercise}
A team's MEMS oscillator passes ageing qualification at $25\,^\circ\text{C}$ but shows accelerated frequency drift at $85\,^\circ\text{C}$ in extended life testing. State the likely failure mechanisms (charge trapping, hermeticity loss, material relaxation) and the corresponding engineering actions.
\end{exercise}
