\chapter{Communications and digital infrastructure}
\label{vol12:ch08}

\epigraph{The internet is a remarkable success of engineering, a remarkable failure of governance, and a remarkable revelation of what happens when the two are separated.}{Author}

\chapmeta{Half-life: medium-life. Archetypes: transport, scaling, ethics. Exercise target: 20. Project track: simulation (assess one digital service's planetary-scale failure modes). Hazard class: high (single-points-of-failure with civilization-scale dependencies). Reader path default: standard, with the internet and submarine-cable sections core and the algorithmic-governance section mastery. Named cases: Facebook BGP outage 2021 (\cite{paper:janardhan-facebook-outage-2021}); Egypt internet shutdown 2011 (\cite{paper:dainotti-egypt-2011}); AWS US-East-1 multiple outages (\cite{web:aws-us-east-1-postmortems}).}

\noindent
The internet is the largest engineered system humans have built. Approximately $5 \times 10^9$ devices connected; approximately $5 \times 10^9$ users; approximately $300 \times 10^{18}$ bytes of annual traffic; approximately $4 \times 10^9$ km of fibre-optic cable (terrestrial and submarine); approximately $10^7$ servers in hyperscale data centres; approximately $10^{10}$ smartphones; and the entire planetary economy increasingly dependent on it operating. The system is the result of approximately $50$ years of engineering work since the ARPANET in 1969 and approximately $30$ years of mass deployment since the World Wide Web in 1991. The chapter takes the internet and its sub-systems (submarine cables; backbone networks; cellular access; satellite communications; data centres; content-distribution networks) as the engineering object, and treats the algorithmic-governance question that emerges at civilization scale.

\section{Communications as civilization infrastructure}\pathtag{core}

Communications became civilization infrastructure incrementally over $150$ years. The telegraph (Morse 1844, Atlantic 1858/1866) was the first instantaneous long-distance communication; the telephone (Bell 1876) extended it to voice; the radio and television (early 20th century) extended it to broadcast; the satellite (Telstar 1962) extended it to global coverage; the internet (ARPANET 1969 then commercial 1990s) extended it to packet-switched data; the smartphone (iPhone 2007 then global rollout) extended it to per-person continuous connectivity. Each layer added scale, function, and dependence; each became, by some interval, civilization infrastructure that the prior society organised itself around.

The contemporary engineering profession works on the next layers: $5$G cellular and the emerging $6$G; high-throughput low-Earth-orbit satellite constellations (Starlink at $\approx 6{,}000$ satellites and expanding, OneWeb, Project Kuiper, Chinese Guowang); the hyperscale data centres that host the cloud; the AI inference infrastructure that the contemporary AI deployment depends on. The pattern: each layer is built on the previous, with the older layers persisting alongside (the telephone POTS network is still operational in most jurisdictions; the IPv4 internet remains the dominant addressing scheme); the new layers add capability without retiring the old.

\begin{archetype}[Scaling: internet traffic]
Global internet traffic doubled approximately every $18$-$24$ months from 1990 through 2010, then continued to double approximately every $36$-$48$ months in the 2010s as the doubling rate slowed. Annual traffic in 2024: $\approx 5\,\text{ZB}$ (zettabytes, $10^{21}$ bytes). Per-user traffic: $\approx 1\,\text{TB}$/yr. The traffic growth has been substantially absorbed through: increasing fibre capacity (dense wavelength-division multiplexing now delivers $>10\,\text{Tb/s}$ per fibre pair); increasing data-centre capacity (hyperscale facilities at $>100\,\text{MW}$ each); increasing content-distribution-network deployment (caches at $>10^4$ locations globally); increasing access-network capacity ($5$G cellular; fibre-to-the-home). The engineering challenge is sustained: each generation of capacity must be planned, built, and deployed against the next generation of demand.
\end{archetype}

\section{Internet at planetary scale}\pathtag{core}

The internet's architecture is layered: physical (fibre, radio, satellite); link (Ethernet, Wi-Fi, cellular); network (IP, principally IPv4 with IPv6 deployment ongoing); transport (TCP, UDP, QUIC); application (HTTP, DNS, SMTP, and a vast portfolio above). The end-to-end principle (Saltzer, Reed, Clark 1981; \cite{paper:saltzer-end-to-end-1984}) is the design philosophy: keep the network simple and put complexity at the endpoints. The result is a network that has scaled by approximately $9$ orders of magnitude (from $\sim 10$ hosts in 1969 to $\sim 10^{10}$ devices in 2024) without architectural revision.

The internet's interconnection structure (the autonomous-system topology; the BGP routing protocol; the IXP/peering ecosystem) is the engineering object that emerged from the commercial transition of the 1990s. Approximately $80{,}000$ autonomous systems globally exchange routing information; approximately $1{,}000$ major IXPs (internet exchange points) host the peering interconnections; the largest few-hundred networks (Tier-1 transit; major content; major cloud) carry the substantial majority of traffic. The architecture is decentralised in principle and concentrated in practice; the engineering implication is that single-point-of-failure analysis is non-trivial.

\begin{principle}[End-to-end principle]
The end-to-end principle: keep the network layers simple and put complexity at the endpoints. The principle enabled the internet to scale across decades of capacity and capability growth by allowing endpoint innovation to proceed without coordinated network-side updates. The principle is challenged by contemporary engineering pressures (in-network caching; deep packet inspection; carrier-grade NAT; CDN architectures that move computation toward the edges) and by the algorithmic-governance pressures the chapter discusses. The principle is not absolute; the engineering profession's judgement is the appropriate balance between the architectural simplicity it enables and the operational and policy requirements it complicates.
\end{principle}

\begin{estimation}
\textbf{Question.} Estimate the global data-centre electricity demand as a fraction of global electricity in 2024 and the projected demand by 2030 under continued AI-inference and cloud-services growth.

\textbf{Reasoning.} Global data-centre electricity demand 2024: $\approx 500\,\text{TWh/yr}$ (IEA estimate; including cryptocurrency and AI). Global electricity: $\approx 30{,}000\,\text{TWh/yr}$. Fraction: $\approx 1.7\%$. Projected 2030 under typical scenarios: $\approx 1{,}000$-$1{,}500\,\text{TWh/yr}$ depending on AI deployment trajectory; fraction $\approx 3\%$-$5\%$. Conclusion: data-centre electricity demand is becoming a significant fraction of global electricity, and the location decisions for data-centre placement are increasingly constrained by grid capacity and water-availability (cooling) in addition to the historical considerations of latency and connectivity.
\end{estimation}

\section{Submarine cables and the actual internet}\pathtag{core}

The internet's physical reality is approximately $500$ submarine cables totalling approximately $1.5 \times 10^6\,\text{km}$ that carry $>95\%$ of international traffic. The cables are owned by consortia of telecommunications carriers, with the contemporary additions of hyperscale operators (Google, Meta, Microsoft, Amazon) who increasingly invest in cable capacity for their own traffic. The capacity per cable: $\approx 200\,\text{Tb/s}$ for current designs, with successive generations doubling roughly every $5$ years through DWDM and coherent-optics improvements.

The submarine-cable engineering combines: the cable itself (optical fibres in a protective armoured cable, with high-voltage power conductors for the repeaters); the repeaters (every $50$-$100\,\text{km}$, amplifying the optical signal); the cable-landing stations on shore; the routing-and-redundancy planning to avoid single-point-of-failure on any given route. The cables are vulnerable to: fishing-trawler damage (the dominant cause of cuts in shallow water); anchor damage; earthquakes (the 2006 Hengchun earthquake off Taiwan severed multiple cables and disrupted Asian connectivity for weeks); deliberate damage (the 2023 Red Sea cable cuts in disputed circumstances). The engineering response is route diversity and rapid repair; the institutional response is the international cable protection agreements and the maritime regulatory framework.

The cable map's geography concentrates traffic on a few choke-points (the Suez/Red Sea corridor; the Strait of Malacca; the Atlantic crossings around Land's End and Brittany; the trans-Pacific corridors), making those corridors strategic infrastructure. The engineering and policy implications of this concentration are an active area of contemporary work.

\section{Cellular networks}\pathtag{standard}

Cellular networks deliver the access tier for mobile communication. The technology trajectory: $1$G (analog voice; 1980s), $2$G (digital voice and SMS; GSM and CDMA; 1990s), $3$G (initial mobile data; 2000s), $4$G/LTE (broadband mobile data; 2010s), $5$G (high-bandwidth, low-latency, machine-type communications; 2020s). Each generation has approximately doubled the achievable data rate per spectrum and expanded the application portfolio.

The $5$G engineering combines three deployment tiers: low-band ($<1\,\text{GHz}$; long range, low capacity); mid-band ($1$-$6\,\text{GHz}$; balanced; the dominant deployment); millimetre-wave ($>24\,\text{GHz}$; very high capacity, very short range; deployment limited to dense urban hot-spots). The base-station density required for $5$G mid-band coverage is approximately the same as $4$G; for millimetre-wave coverage, $10$-$100\times$ denser. The infrastructure capital investment for full national $5$G coverage in a typical developed country: $\$20$-$\$50$ billion over $5$-$10$ years.

The mobile network's planetary-scale role: approximately $5 \times 10^9$ mobile-broadband subscriptions globally; approximately $90\%$ of the global population has $4$G or better coverage; approximately $5\%$ of the global population remains unserved by mobile broadband. The engineering and policy question for the next decade: the closing of the access gap in low-income regions; the role of satellite-direct-to-handset services (Starlink-direct, AST SpaceMobile) in the closure.

\section{Satellites and orbital infrastructure}\pathtag{standard}

Satellite communications have evolved from the geostationary-satellite era (Telstar 1962 through the early 2000s; high latency, modest capacity, dominant for broadcast TV and long-distance voice) to the contemporary low-Earth-orbit constellations (Iridium 1998-2001 as early entrant; Starlink since 2019 as the breakthrough). The low-Earth-orbit advantages: latency comparable to terrestrial fibre ($20$-$50\,\text{ms}$); much higher capacity per satellite; per-user costs competitive with mobile broadband. The disadvantages: many more satellites required (Starlink at $\approx 6{,}000$ as of 2024; projected to grow to $\sim 12{,}000$; competitors building constellations of $\sim 1{,}000$-$10{,}000$ each); orbital-debris and conjunction-management implications; substantial concentration of capability in a few operators.

The contemporary engineering and policy questions: orbital-traffic management (the increase from $\sim 2{,}000$ active satellites in 2019 to $\sim 10{,}000$ in 2024 to $\sim 50{,}000$ projected for 2030 stresses the orbital-debris environment; Volume XII Chapter 10 elaborates); spectrum allocation (the V/E/W-band allocations that the constellations require); the equity question of whether the constellations serve global connectivity or principally extract economic value to home-country operators; the strategic implications of the concentration. Volume XII Chapter 10 takes space infrastructure as its principal topic.

\section{Algorithmic governance: who controls flow}\pathtag{mastery}

The internet's algorithmic-governance question emerged with the consolidation of traffic across a few hyperscale platforms over the 2010s. Approximately $50\%$ of global internet traffic now traverses a few-hundred companies' networks and platforms; the algorithmic decisions made within those companies (search ranking; social-media feed ordering; advertisement targeting; content moderation; recommendation algorithms) determine substantial aspects of the contemporary public sphere. The engineering profession's responsibility includes the design choices for these algorithms; the institutional and political-economic question is whether the responsibility is being exercised with appropriate accountability to the populations affected.

The contemporary debate (engineering, institutional, regulatory) includes: content-moderation at scale (the trade-offs among free expression, harmful content, regulatory compliance, and operational cost); recommendation-algorithm effects on information diet (the contested empirical question of recommendation-driven polarisation); advertising-and-tracking architecture (the privacy-and-monitoring implications of the surveillance-advertising business model); and the wider governance question of platform power. The engineering profession's contribution is the technical analysis and the design alternatives; the political-economic and ethical questions are appropriately broader than the engineering scope.

\section{Failure: undersea cable cuts, BGP at planetary scale, content-moderation as engineering}\pathtag{core}

The Facebook BGP outage of 4 October 2021 \cite{paper:janardhan-facebook-outage-2021} is the contemporary canonical case of digital-infrastructure single-point-of-failure: a BGP misconfiguration during routine maintenance withdrew Facebook's routes from the global routing table; the entire Facebook portfolio (Facebook, Instagram, WhatsApp, Oculus, internal Facebook services) became unreachable; the engineering team that needed to fix the problem was unable to log in to the building because the badge-access system also depended on Facebook-internal services; the recovery took approximately $6$ hours. The technical mechanism: BGP and DNS dependencies that the architecture had not adequately separated. The civilizational implications: approximately $3 \times 10^9$ users were affected; secondary effects included spikes in load on competing services; the outage demonstrated how concentrated essential communications had become.

The 2011 Egypt internet shutdown \cite{paper:dainotti-egypt-2011} is the case of state-controlled internet flow: the Egyptian government instructed the country's BGP-speaking ISPs to withdraw the country's IP address space from the global routing table; the result was the disconnection of approximately $90\%$ of Egypt's internet for approximately five days. The technical mechanism: BGP route withdrawal. The institutional implication: the internet's distributed routing system is decentralised in principle and locally controllable in practice; the engineering response to the contemporary internet-shutdown regime (China; North Korea; Russia post-2022; periodic in many other jurisdictions) is the limitation of the addressing-and-routing-control concentration.

The AWS US-East-1 region's series of outages \cite{web:aws-us-east-1-postmortems} (notably 2017, 2020, 2021) are the case of cloud-infrastructure concentration risk: the region hosts a substantial share of internet services for many companies that have nominal multi-region architectures but actual cross-region dependencies; outages in the region cascade through dozens of major services for hours. The technical mechanism varies; the structural problem is the cross-service-dependency concentration. The engineering response: explicit cross-region dependency analysis; chaos-engineering practices that test for the concentrations; multi-cloud architectures for the most critical applications.

\begin{principle}[Decentralisation as engineering scope]
The internet's intended decentralisation was a design choice motivated by survivability under adversarial conditions; the contemporary deployment has substantially concentrated the operational reality in a few platforms, networks, and data-centres. The chapter's failure cases each illustrate the consequences of the concentration. The engineering response: explicit decentralisation as a design objective (multi-region architectures; multi-provider dependencies; route diversity); regulatory frameworks that constrain the concentration; standards and protocols that support distributed alternatives. The engineering profession's responsibility includes the conscious choice between concentrated and decentralised architectures and the trade-offs each entails.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Digital-service planetary-scale failure analysis]
Choose a widely-used digital service (a social-media platform; a payment-processing service; a major SaaS application; a streaming platform). Document the service's architectural dependencies: cloud providers; CDN providers; DNS providers; submarine-cable corridors; data-centre regions; key third-party services. Identify the planetary-scale failure modes (regional outages; cross-service dependencies; routing-attack vulnerabilities; supply-chain compromises). Estimate the user-and-economic impact of each failure mode; identify the engineering interventions that would reduce the failure probability or consequence; identify the institutional and contractual structures that would support the engineering interventions.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A submarine cable carries $200\,\text{Tb/s}$ across a $10{,}000\,\text{km}$ span with repeaters every $80\,\text{km}$. Compute the number of repeaters and the total cable power consumption at $1\,\text{W}$/repeater.
\end{exercise}

\begin{exercise}
A hyperscale data centre operates at $100\,\text{MW}$ with $1.2$ PUE (power-usage effectiveness). Compute the IT load, the cooling-and-overhead load, the annual electricity consumption, and the cost at \$0.05/kWh.
\end{exercise}

\begin{exercise}
A CDN node serves $10{,}000$ requests/s at $100\,\text{KB}$ average response size. Compute the aggregate egress bandwidth and the upstream-internet bandwidth required.
\end{exercise}

\begin{exercise}
A $5$G base station has $100\,\text{m}$ outdoor coverage radius in millimetre-wave. Compute the number of base stations required to cover a $10\,\text{km}^2$ urban area and compare to the $1{,}000\,\text{m}$ coverage of a mid-band $5$G site.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the optimal redundancy strategy for a critical service across multiple cloud regions, given regional-outage probability, failover time, and cross-region traffic cost.
\end{exercise}

\begin{exercise}
Derive the cost-and-latency trade-off between centralised data-centre architecture and edge-computing distributed architecture for a typical workload.
\end{exercise}

\begin{exercise}
Derive the Shannon capacity bound for a fibre-optic link and identify how DWDM achieves throughput substantially above the per-wavelength Shannon limit.
\end{exercise}

\begin{exercise}
Derive the orbital-mechanics constraints on a LEO constellation: number of satellites required for global coverage at given orbital altitude and inclination.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the global data-centre water consumption for cooling and project the implications under continued data-centre expansion.
\end{exercise}

\begin{exercise}
Estimate the strategic-concentration metric for global internet infrastructure: the fraction of total traffic that traverses the top $10$ networks, the top $10$ submarine cables, and the top $10$ data centres.
\end{exercise}

\begin{exercise}
Estimate the embedded carbon of the global internet (manufacturing and disposal of routers, switches, fibres, base stations, end-user devices).
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate BGP-propagation dynamics under a major route-hijacking attack. Identify the network configurations that limit propagation versus those that allow rapid spread.
\end{exercise}

\begin{exercise}
Simulate a multi-region cloud-application failure scenario. Identify the dependency configurations that contain the failure versus those that produce cascading outage.
\end{exercise}

\begin{exercise}
Simulate the load on competing services during a major platform outage. Identify the capacity-headroom required to absorb the spike.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a multi-region cloud-application architecture with explicit cross-region dependency analysis. Specify the data-replication strategy, the failover mechanisms, the operational procedures.
\end{exercise}

\begin{exercise}
Design a submarine-cable consortium for a new trans-Atlantic cable. Specify the cable parameters, the landing-station locations, the ownership and capacity allocation, the route-redundancy strategy.
\end{exercise}

\begin{exercise}
Design an internet-shutdown-resistant communication architecture for civil-society organisations operating in jurisdictions with periodic internet shutdowns. Specify the technical and operational elements.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Discuss the Facebook BGP outage of October 2021. Identify the engineering and operational decisions that produced the failure; identify the structural concentrations that determined its impact; identify the engineering interventions that would prevent recurrence.
\end{exercise}

\begin{exercise}
Argue: the contemporary internet's concentration in a few hyperscale platforms is a market failure that engineering alone cannot solve; regulatory intervention is required. Argue the opposite. Identify the engineering implications of each position.
\end{exercise}

\begin{exercise}
Discuss the engineering profession's role in algorithmic-governance design. Identify the technical contributions; identify the institutional and political-economic limits.
\end{exercise}
