\chapter{Statistical inference}
\label{vol02:ch14}

\epigraph{All models are wrong, but some are useful.}{George
E. P. Box, working aphorism in the time-series and quality-control
literature; the closest published form is in Box, ``Robustness in
the strategy of scientific model building,'' 1979.}

\chapmeta{Half-life: foundational. Archetypes: uncertainty
(estimator uncertainty as a first-class quantity), ethics
(statistical malpractice as a failure mode the engineer must name).
Exercise target: 40. Project track: simulation with analysis (a
linear, logistic, or regularised regression on a real dataset, with
$k$-fold cross-validation and an honest residual diagnostic, plus a
1500-word reflection). Hazard class: none. Reader path default:
standard, with sections explicitly tagged. Named cases: none in this
chapter. The replication-crisis discussion in section 14.7 cites
five primary methods papers and is not registered under named
accidents; it is methodological evidence, not a single event.}

\noindent
Volume I Chapter 8 introduced the engineer to summary statistics,
five working distributions, the $p$-value, the confidence and
prediction intervals, the Bayesian contingency table, and a first
catalogue of statistical malpractice. The mathematics there was
applied on trust. This chapter supplies the formal apparatus those
formulas presupposed: the estimator as a function of data, its bias
and variance, the likelihood function, the Neyman-Pearson framing
of testing, the Bayesian update as a posterior density, the
regression model as a projection plus a noise structure, and
cross-validation as the discipline that prevents an engineer from
reporting a number their data has already chosen. The malpractice
catalogue of Volume I Chapter 8 returns at the end with primary
citations, formal definitions, and the multiple-comparisons
correction.

\begin{archetype}[Uncertainty, promoted to a first-class object]
Volume I carried uncertainty as a single number $u(x)$ beside each
measurement, and Volume I Chapter 8 traced $u(\bar{x}) = s/\sqrt{n}$
to variation across a sample. This chapter makes the estimator
itself the object of study. Every quantity the engineer reports from
data, a mean, a slope, a defect rate, a regularisation parameter, is
a function of a random sample and therefore carries its own
distribution. The bias, the variance, the sampling distribution, and
the coverage of an interval are properties of that distribution, and
the engineer who reports a number without its distribution has
reported half a result. The chapter's discipline is to attach the
distribution to every estimate.
\end{archetype}

\begin{archetype}[Ethics: malpractice as a named failure mode]
The second archetype is the one Volume I Chapter 8 introduced and
this chapter formalises. A statistical procedure can be technically
correct at every step and still produce a number that misrepresents
the evidence, because the choice of which procedure to run was made
after the data were seen, or because the number of tests run was not
disclosed. The failure is not in the arithmetic. It is in the
reporting discipline. Section 14.7 names the mechanisms, attaches the
controlling correction to each, and connects the catalogue to the
replication studies that have measured how large the problem is in
published literatures the engineer might draw on for design input.
\end{archetype}

\section{Estimation: point and interval}\pathtag{core}

A \engterm{statistical model} is a family of probability
distributions, indexed by a parameter $\theta$ in a parameter set
$\Theta$. We write the family as $\set*{p(x; \theta) : \theta \in
\Theta}$ for densities or as $\set*{P_{\theta} : \theta \in
\Theta}$ for distributions. The data $X_{1}, \ldots, X_{n}$ are
modelled as independent draws from one member of the family, and the
engineer's task is to identify which member. Volume II Chapter 13
provides the probability backbone; this chapter takes the random
variables for granted and works on the parameter.

A \engterm{statistic} is any function of the data, $T = T(X_{1},
\ldots, X_{n})$. A statistic that is intended as a guess for
$\theta$ is an \engterm{estimator}, written $\hat{\theta} =
\hat{\theta}(X_{1}, \ldots, X_{n})$. The hat is notation, not a
licence: an estimator is a random variable, with its own
distribution induced by the distribution of the data
\cite[ch.~6]{text:wasserman2004}.

\subsection{Bias, variance, mean squared error}

The two scalar summaries that govern an estimator's quality are its
bias and its variance. The \engterm{bias} of $\hat{\theta}$ at the
true parameter $\theta$ is
\[
\operatorname{bias}_{\theta}(\hat{\theta}) = \E_{\theta}[\hat{\theta}] - \theta.
\]
The \engterm{variance} is $\Var_{\theta}(\hat{\theta}) =
\E_{\theta}[(\hat{\theta} - \E_{\theta}[\hat{\theta}])^{2}]$.
The \engterm{mean squared error} combines both,
\[
\operatorname{MSE}_{\theta}(\hat{\theta})
= \E_{\theta}[(\hat{\theta} - \theta)^{2}]
= \Var_{\theta}(\hat{\theta}) + \operatorname{bias}_{\theta}(\hat{\theta})^{2}.
\]
The bias-variance decomposition is the operational identity behind
every later choice in this chapter, including ridge regression,
cross-validation, and Bayesian shrinkage. An estimator with smaller
MSE than another may still be more biased: the variance reduction
can outweigh the bias penalty. This is the engineer's licence to
prefer a slightly biased estimator when its variance is much
smaller. We come back to this trade-off formally in section 14.5.

\paragraph{Worked example: two estimators of a uniform endpoint.}
A sensor reports values modelled as $\mathrm{Uniform}[0, \theta]$,
and the engineer wants $\theta$. Two estimators compete. The
first doubles the sample mean, $\hat{\theta}_{1} = 2\bar{X}_{n}$,
on the reasoning that $\E[X] = \theta/2$. The second takes the
sample maximum, $\hat{\theta}_{2} = \max_{i} X_{i}$. We compare
their bias and variance directly. For the moment estimator,
$\E[\hat{\theta}_{1}] = 2\E[\bar{X}_{n}] = 2 \cdot \theta/2 =
\theta$, so $\hat{\theta}_{1}$ is unbiased; its variance is
$\Var(2\bar{X}_{n}) = 4\Var(\bar{X}_{n}) = 4 \cdot
(\theta^{2}/12)/n = \theta^{2}/(3n)$, using
$\Var(X) = \theta^{2}/12$ for the uniform. For the maximum, the
cdf of $\hat{\theta}_{2}$ is $(t/\theta)^{n}$ on $[0, \theta]$,
giving $\E[\hat{\theta}_{2}] = \frac{n}{n+1}\theta$ and
$\E[\hat{\theta}_{2}^{2}] = \frac{n}{n+2}\theta^{2}$. The maximum
is biased low by $-\theta/(n+1)$, and its variance is
$\frac{n}{n+2}\theta^{2} - \bigl(\frac{n}{n+1}\bigr)^{2}\theta^{2}
= \frac{n\,\theta^{2}}{(n+1)^{2}(n+2)}$. The mean squared errors
are
\[
\operatorname{MSE}(\hat{\theta}_{1}) = \frac{\theta^{2}}{3n},
\qquad
\operatorname{MSE}(\hat{\theta}_{2})
= \frac{n\,\theta^{2}}{(n+1)^{2}(n+2)} + \frac{\theta^{2}}{(n+1)^{2}}
= \frac{2\theta^{2}}{(n+1)(n+2)}.
\]
At $n = 10$, $\operatorname{MSE}(\hat{\theta}_{1}) = \theta^{2}/30
\approx 0.0333\,\theta^{2}$ and $\operatorname{MSE}(\hat{\theta}_{2})
= 2\theta^{2}/132 \approx 0.0152\,\theta^{2}$. The biased estimator
wins by more than a factor of two, and its advantage grows as
$1/n^{2}$ against the moment estimator's $1/n$. The bias-corrected
maximum $\frac{n+1}{n}\hat{\theta}_{2}$ removes the bias and lowers
the MSE further, to $\theta^{2}/(n(n+2))$. The example is the
chapter's first demonstration that unbiasedness is not the
objective; mean squared error is.

\paragraph{Worked example: the bias of the plug-in variance.}
The maximum-likelihood variance $\hat{\sigma}^{2}_{\mathrm{MLE}}
= \frac{1}{n}\sum(X_{i} - \bar{X})^{2}$ has expectation
$\frac{n-1}{n}\sigma^{2}$, so its bias is $-\sigma^{2}/n$ and its
squared bias is $\sigma^{4}/n^{2}$. Under a Gaussian model its
variance is $2(n-1)\sigma^{4}/n^{2}$, giving
$\operatorname{MSE}(\hat{\sigma}^{2}_{\mathrm{MLE}}) =
\frac{2n-1}{n^{2}}\sigma^{4}$. The Bessel-corrected estimator
$S_{n}^{2}$ is unbiased but has variance $2\sigma^{4}/(n-1)$,
giving $\operatorname{MSE}(S_{n}^{2}) = 2\sigma^{4}/(n-1)$. For
every $n \geq 2$ the biased MLE has the smaller mean squared error.
The estimator $\frac{1}{n+1}\sum(X_{i}-\bar{X})^{2}$, more biased
still, minimises MSE among scalar multiples of the sum of squares
under the Gaussian model. The reader who took the $n-1$ denominator
as the right answer in Volume I now sees that ``right'' depends on
the loss: $n-1$ minimises bias, $n+1$ minimises mean squared error,
and $n$ is the maximum-likelihood compromise.

Figure \ref{fig:vol02:ch14:mse-tradeoff-uniform} draws the three
uniform-endpoint estimators together, in units of $\theta^{2}$ against
sample size, and the picture carries the lesson that the two worked
examples argued in symbols: the biased estimators sit below the
unbiased one at every sample size, and the gap widens because their
error falls as $1/n^{2}$ against the moment estimator's $1/n$.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-mse-tradeoff-uniform.tex}

\paragraph{Worked example: a gauge study's variance components.} The
bias-variance decomposition also governs how an engineer reads a
measurement-system study, where the observed variance of a set of
gauged parts splits into the true part-to-part variance and the gauge's
own measurement variance, $\sigma_{\mathrm{obs}}^{2} =
\sigma_{\mathrm{part}}^{2} + \sigma_{\mathrm{gauge}}^{2}$. A study
measures ten parts twice each and returns an observed variance of
$\hat{\sigma}_{\mathrm{obs}}^{2} = 4.00$ working units and a
repeatability variance, estimated from the paired remeasurements, of
$\hat{\sigma}_{\mathrm{gauge}}^{2} = 0.36$. The part variance is
recovered by subtraction, $\hat{\sigma}_{\mathrm{part}}^{2} = 4.00 -
0.36 = 3.64$, and the \engterm{gauge capability ratio} is
$\hat{\sigma}_{\mathrm{gauge}}/\hat{\sigma}_{\mathrm{obs}} =
0.60/2.00 = 0.30$, so the gauge contributes thirty percent of the
observed spread, above the ten-percent working target that a capable
gauge should meet. The subtraction is the operational reason the
engineer never reports the gauge variance as if it were the process
variance: the two are additive, and a study that ignores the gauge
component reports a process more variable than it is. The estimator
$\hat{\sigma}_{\mathrm{part}}^{2}$ is a difference of two variance
estimates, so its own variance is the sum of their variances, and a
small part variance recovered as a difference of two large ones is a
poorly determined number, which is why a gauge study sizes its
replication to keep $\hat{\sigma}_{\mathrm{gauge}}^{2}$ small relative
to $\hat{\sigma}_{\mathrm{part}}^{2}$ before the subtraction is trusted.

\subsection{Method of moments}

The \engterm{method of moments} is the oldest systematic estimator
and the engineer's fastest route to a starting value. The
population moments $\mu_{k} = \E[X^{k}]$ are functions of the
parameter $\theta$. The method sets the first $d$ population
moments equal to the first $d$ sample moments $\hat{\mu}_{k} =
\frac{1}{n}\sum_{i} X_{i}^{k}$ and solves the resulting system for
$\theta$ \cite[ch.~9]{text:wasserman2004}.

\paragraph{Worked example: a Gamma shape and rate.}
For $X_{i} \sim \mathrm{Gamma}(\alpha, \beta)$ with shape $\alpha$
and rate $\beta$, the population mean is $\alpha/\beta$ and the
population variance is $\alpha/\beta^{2}$. Matching the first two
moments, $\bar{X} = \alpha/\beta$ and the sample variance
$\hat{\mu}_{2} - \bar{X}^{2} = \alpha/\beta^{2}$, divides to give
$\hat{\beta} = \bar{X}/(\hat{\mu}_{2} - \bar{X}^{2})$ and
$\hat{\alpha} = \bar{X}\,\hat{\beta} = \bar{X}^{2}/(\hat{\mu}_{2} -
\bar{X}^{2})$. With a sample mean of $4.0$ and a sample variance
of $2.0$, the method gives $\hat{\beta} = 2.0$ and $\hat{\alpha}
= 8.0$. The Gamma likelihood has no closed-form MLE (it requires
solving a transcendental equation in the digamma function), so the
method-of-moments value is the standard initialiser for the
Newton iteration that finds the MLE. The method of moments is
consistent under mild conditions but is generally less efficient
than the MLE; its working role is as a fast starting point for the
iteration that finds the MLE.

\subsection{Consistency and asymptotic normality}

An estimator $\hat{\theta}_{n}$, computed from a sample of size $n$,
is \engterm{consistent} for $\theta$ if $\hat{\theta}_{n}
\xrightarrow{P} \theta$ as $n \to \infty$, in the sense of
convergence in probability \cite[ch.~5, ch.~6]{text:wasserman2004}.
Consistency is the minimum guarantee a working estimator should
satisfy; an inconsistent estimator does not improve with more data.

An estimator is \engterm{asymptotically normal} if
\[
\sqrt{n}\,(\hat{\theta}_{n} - \theta)
\xrightarrow{d} \mathcal{N}\!\left(0,\, V(\theta)\right),
\]
for some asymptotic variance $V(\theta)$. The maximum-likelihood
estimator and the least-squares estimator both satisfy this under
standard regularity conditions, and the asymptotic-normality
property is what justifies the textbook habit of reporting an
estimate as $\hat{\theta} \pm z_{1-\alpha/2}\, \widehat{\mathrm{se}}(\hat{\theta})$
when $n$ is large.

The sampling distribution is the object that makes ``$\hat{\theta}$
is a random variable'' concrete. Figure
\ref{fig:vol02:ch14:sampling-dist} shows the sampling distribution
of the sample mean narrowing as $n$ grows: the spread is the
standard error $\sigma/\sqrt{n}$, which shrinks as $1/\sqrt{n}$, and
the shape approaches the Gaussian by the central limit theorem of
Volume II Chapter 13 even when the underlying data are not Gaussian.
The figure is the picture behind every confidence interval in this
chapter.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-sampling-dist.tex}

\subsection{The sample mean and sample variance}

The \engterm{sample mean} $\bar{X}_{n} = \frac{1}{n}\sum_{i=1}^{n}
X_{i}$ is unbiased for the population mean $\mu = \E[X_{1}]$ and
has variance $\sigma^{2}/n$ when the data are independent and
identically distributed with population variance $\sigma^{2}$.
The \engterm{sample variance} with the Bessel correction,
\[
S_{n}^{2} = \frac{1}{n-1}\sum_{i=1}^{n} (X_{i} - \bar{X}_{n})^{2},
\]
is unbiased for $\sigma^{2}$. The denominator $n-1$, rather than
$n$, is the correction the reader took on trust in Volume I; the
algebra exercise at the end of this chapter shows it.

\subsection{The plug-in estimator}

For a scalar functional $T(F)$ of an unknown distribution $F$, the
\engterm{plug-in estimator} substitutes the empirical distribution
$\hat{F}_{n}$, the distribution that places mass $1/n$ at each
observation, into the functional:
\[
\hat{T}_{n} = T(\hat{F}_{n}).
\]
The plug-in mean recovers the sample mean. The plug-in variance
recovers the maximum-likelihood (uncorrected) variance estimate.
The plug-in approach is the engineer's first line of attack on any
functional that does not have a tabled estimator.

\paragraph{Worked example: a plug-in skewness estimate.} The skewness
$\gamma = \E[(X - \mu)^{3}]/\sigma^{3}$ has no textbook standard-error
formula, so the plug-in estimator is the natural first reach:
substitute the empirical distribution and read off
$\hat{\gamma} = \frac{1}{n}\sum_{i}(x_{i} - \bar{x})^{3} /
\bigl(\frac{1}{n}\sum_{i}(x_{i} - \bar{x})^{2}\bigr)^{3/2}$. Six
measured values $\set{2, 3, 3, 4, 5, 13}$ have $\bar{x} = 5$; the
centred moments are $\frac{1}{6}\sum(x_{i}-5)^{2} = (9 + 4 + 4 + 1 + 0
+ 64)/6 = 13.67$ and $\frac{1}{6}\sum(x_{i}-5)^{3} = (-27 - 8 - 8 - 1
+ 0 + 512)/6 = 76.67$, so $\hat{\gamma} = 76.67/13.67^{3/2} =
76.67/50.53 = 1.52$, a strong right skew driven by the single value at
$13$. The plug-in estimate is biased for small $n$ (the sample third
moment underestimates the population third moment much as the plug-in
variance underestimates $\sigma^{2}$), and it carries no closed-form
standard error, which makes it the natural first client of the
bootstrap the next subsection introduces: resampling supplies the
standard error the plug-in functional does not.

\subsection{Confidence intervals: definition and construction}

A \engterm{$1-\alpha$ confidence interval} for $\theta$ is a pair of
data-dependent endpoints $L(X), U(X)$ such that
\[
\Pr_{\theta}\!\left[L(X) \leq \theta \leq U(X)\right] \geq 1-\alpha
\quad \text{for all } \theta \in \Theta.
\]
The probability is over the data, with $\theta$ fixed. The interval
is random; the parameter is not. The procedure has $1-\alpha$
long-run coverage. The frequentist statement ``the interval covers
$\theta$ with probability $0.95$'' refers to the procedure across
hypothetical resamples, not to a posterior belief about a particular
realised interval.

The standard pivot construction works as follows. If
$\sqrt{n}(\hat{\theta} - \theta)/\widehat{\mathrm{se}}(\hat{\theta})$
has an approximately standard normal distribution at sample size
$n$, then
\[
\hat{\theta} \pm z_{1-\alpha/2}\,\widehat{\mathrm{se}}(\hat{\theta})
\]
is a $1-\alpha$ confidence interval. With unknown $\sigma$ and a
Gaussian model, $z_{1-\alpha/2}$ is replaced by the $t$-quantile
$t_{1-\alpha/2,n-1}$, recovering the Volume I formula.

\paragraph{The coverage statement, made operational.} The single
most common misreading of a confidence interval is to read the
$95\,\%$ as a probability about the realised interval. The realised
interval $[11.5, 13.3]$ either contains $\theta$ or it does not;
there is no probability left once the data are in hand. The
$95\,\%$ describes the \emph{procedure}: across many independent
datasets, $95\,\%$ of the intervals the procedure produces will
contain the fixed true $\theta$. Figure
\ref{fig:vol02:ch14:ci-coverage} draws fifty intervals from fifty
simulated samples; the intervals that miss the true value are
drawn in red, and the long-run miss rate is the $\alpha = 0.05$ the
construction promises. The misreading matters in practice because
it licenses the engineer to attach a posterior-sounding confidence
to one number, which is the Bayesian credible interval's job, not
the frequentist confidence interval's.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-ci-coverage.tex}

\paragraph{Worked example: a finite-sample coverage check.}
A $90\,\%$ confidence interval for a Gaussian mean at $n = 5$ uses
$t_{0.95, 4} = 2.132$, not the asymptotic $z_{0.95} = 1.645$. The
difference is the small-sample penalty: the interval is wider by a
factor $2.132/1.645 = 1.30$, a $30\,\%$ inflation that the engineer
who reaches for $z$ at $n = 5$ would silently omit, producing an
interval with true coverage near $0.83$ rather than $0.90$. The
$t$-correction is the price of estimating $\sigma$ from the same
five points. As $n$ grows the penalty shrinks: at $n = 30$ the
$t$-to-$z$ ratio is $1.70/1.645 = 1.03$, a $3\,\%$ inflation, and
by $n = 120$ it is under $1\,\%$. The working rule the chapter
endorses is to use $t$ always; the cost is negligible at large $n$
and material at small $n$.

\paragraph{Worked example: Wald against Wilson for a proportion near
the boundary.} The confidence interval for a binomial proportion is
where the pivot construction most often misleads, because the standard
Wald interval $\hat{p} \pm z\sqrt{\hat{p}(1-\hat{p})/n}$ uses a
standard error that collapses to zero as $\hat{p}$ approaches zero or
one. A reliability test observes $s = 1$ failure in $n = 20$ trials,
so $\hat{p} = 0.05$. The Wald $95\,\%$ interval is $0.05 \pm 1.96
\sqrt{0.05 \cdot 0.95/20} = 0.05 \pm 1.96 \cdot 0.0487 = [-0.045,
0.145]$, whose lower endpoint is a negative probability, an interval
the parameter space forbids. The \engterm{Wilson interval} inverts the
score test instead of the Wald test and stays inside $[0, 1]$ by
construction; its endpoints are
\[
\frac{\hat{p} + \frac{z^{2}}{2n} \pm z\sqrt{\frac{\hat{p}(1-\hat{p})}{n}
+ \frac{z^{2}}{4n^{2}}}}{1 + \frac{z^{2}}{n}},
\]
which for these numbers evaluate to
$\frac{0.05 + 0.096 \pm 1.96\sqrt{0.002375 + 0.0024}}{1.192} =
\frac{0.146 \pm 1.96 \cdot 0.0690}{1.192} = \frac{0.146 \pm 0.135}
{1.192}$, giving $[0.009, 0.236]$, an interval that is inside the
parameter space and asymmetric about $\hat{p}$, wider on the high side
where the evidence is thinner. The Wilson interval's coverage stays
near the nominal $95\,\%$ where the Wald interval's coverage collapses
toward $0.80$ for small $s$, which is the reason the working rule for a
proportion near a boundary is to reach for Wilson or the closely
related Agresti-Coull adjustment, never the bare Wald interval.

\subsection{The bootstrap}

When the estimator has no closed-form standard error and no
asymptotic formula the engineer is willing to trust, the
\engterm{bootstrap} provides a working substitute
\cite{paper:v2ch14-efron-bootstrap-1979}. The idea is the
plug-in principle applied to the sampling distribution. The
empirical distribution $\hat{F}_{n}$ stands in for the unknown
population $F$. Repeated samples of size $n$ are drawn from
$\hat{F}_{n}$ with replacement; the statistic of interest is
recomputed on each resample; the empirical distribution of the
$B$ resample-statistics is the \engterm{bootstrap distribution}
of the estimator. The bootstrap standard error is the standard
deviation of the resample-statistics; the percentile interval
takes the $\alpha/2$ and $1-\alpha/2$ quantiles of the bootstrap
distribution as confidence-interval endpoints. The procedure is
the engineer's default when no asymptotic formula is available.
Its working failures are predictable: the bootstrap is unreliable
for statistics that depend on extremes (sample maxima, sample
minima, quantiles near the support boundary) and for heavy-tailed
distributions where the sample-quantile substitution becomes
unstable \cite{paper:ch08-efron-tibshirani-1993}. The full-page
schematic in figure \ref{fig:vol02:ch14:bootstrap} reads the
procedure off as a single diagram.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-bootstrap.tex}

\paragraph{Worked example: a bootstrap standard error for the
median.} The sample median has no simple closed-form standard
error, which is the situation the bootstrap was built for. Consider
eleven measured pull-test forces (in newtons): $\set{210, 225, 198,
240, 215, 205, 260, 195, 230, 218, 222}$, with sample median
$218$. The engineer draws $B = 2000$ resamples of size $11$ with
replacement, computes the median of each, and takes the standard
deviation of the $2000$ resample medians as the bootstrap standard
error. For data of this spread the bootstrap standard error of the
median lands near $7$ to $9\,$N, larger than the
$\sqrt{\pi/2} \approx 1.25$ inflation a Gaussian theory would apply
to the standard error of the mean ($\hat{\sigma}/\sqrt{n} \approx
19/3.32 \approx 5.7\,$N). The median's larger standard error is the
price of its robustness; the bootstrap quantifies the price without
a distributional assumption. The percentile interval reads the
$2.5$th and $97.5$th percentiles of the resample medians directly,
giving a confidence interval that need not be symmetric about the
point estimate. The script at
\repopath{volumes/02-form/ch14-statistical-inference/code/bootstrap\_demo.py}
runs the computation and is the reference implementation for the
project.

\paragraph{Worked example: a bootstrap interval for a correlation.}
The Pearson correlation is the second natural client of the bootstrap,
because its sampling distribution is skewed and bounded in $[-1, 1]$,
so a symmetric $\hat{r} \pm z\,\widehat{\mathrm{se}}$ interval can spill
past one. Eight paired measurements of a coating thickness and its
adhesion strength give a sample correlation $\hat{r} = 0.82$. The
classical route uses \engterm{Fisher's $z$-transform}, $z =
\operatorname{artanh}(r) = \frac{1}{2}\ln\frac{1+r}{1-r}$, which is
approximately Gaussian with standard error $1/\sqrt{n-3}$: here $z =
\operatorname{artanh}(0.82) = 1.157$, standard error $1/\sqrt{5} =
0.447$, interval $1.157 \pm 1.96 \cdot 0.447 = [0.281, 2.033]$ on the
$z$ scale, back-transformed by $r = \tanh(z)$ to $[0.274, 0.966]$.
The bootstrap reaches the same interval without the transform: resample
the eight pairs with replacement, recompute $\hat{r}$ on each of $B =
2000$ resamples, and read the $2.5$th and $97.5$th percentiles, which
for data this correlated land near $[0.30, 0.96]$, close to the
Fisher interval and asymmetric about $0.82$ in the same direction. The
agreement is the reassurance that the small-sample interval is
trustworthy; a disagreement would have told the engineer that eight
pairs are too few for either approximation and that the correlation
should be reported as a point estimate with an explicit small-sample
caveat rather than a precise interval. The bootstrap of a correlation
also exposes a resampling subtlety the median example did not: the
pairs must be resampled together, as units, because resampling the two
columns independently would destroy the very association the statistic
measures.

\subsection{Confidence versus prediction}

A confidence interval reports uncertainty about a parameter (a
property of the population). A \engterm{prediction interval} reports
uncertainty about a future observation (a property of an individual
draw). The two are distinct quantities and have distinct widths;
Volume I Chapter 8 worked the operational example. We treat the
confidence interval as the chapter's default and refer the reader
to Volume I Chapter 8 for the prediction-interval construction.

\subsection{Tolerance intervals: covering a proportion of the population}

A third interval answers a question the first two do not, and it is
the interval a materials specification most often needs. A confidence
interval brackets the mean; a prediction interval brackets the next
single draw; a \engterm{tolerance interval} brackets a stated
proportion of the whole population, with a stated confidence. A
$95/99$ tolerance interval is the interval that contains at least
$99\,\%$ of the population with $95\,\%$ confidence, and it is the
right object when a drawing calls out that ninety-nine percent of
produced parts must fall within a dimensional band. The distinction
matters because the three intervals have three different limiting
behaviours. The confidence half-width shrinks to zero as $1/\sqrt{n}$,
because enough data pin the mean exactly; the prediction and tolerance
half-widths shrink toward a fixed multiple of the population standard
deviation, because a future draw and a population proportion both
carry irreducible spread that no sample size removes.

For a Gaussian population the two-sided tolerance interval is
$\bar{x} \pm k\,s$, where the \engterm{tolerance factor} $k =
k(n, p, 1-\alpha)$ depends on the sample size $n$, the population
proportion $p$ to be covered, and the confidence $1-\alpha$. The
factor is larger than both the confidence and the prediction
multiplier, because it must simultaneously cover a population fraction
and hedge against the sample's estimate of the mean and standard
deviation being off. A workable closed-form approximation for the
two-sided factor is Howe's,
\[
k \approx z_{(1+p)/2}\,
\sqrt{\frac{(n-1)\bigl(1 + \tfrac{1}{n}\bigr)}
{\chi^{2}_{\alpha,\, n-1}}},
\]
where $z_{(1+p)/2}$ is the standard-normal quantile for the covered
proportion and $\chi^{2}_{\alpha, n-1}$ is the lower-$\alpha$
chi-square quantile that discounts the sample standard deviation for
its own uncertainty.

\paragraph{Worked example: a dimensional tolerance interval.} A run of
$n = 30$ machined pins has sample mean diameter $\bar{x} =
12.00\,\si{\milli\meter}$ and sample standard deviation $s =
0.040\,\si{\milli\meter}$, and the specification asks for the band that
contains $99\,\%$ of pins with $95\,\%$ confidence. The covered
proportion is $p = 0.99$, so $z_{(1+p)/2} = z_{0.995} = 2.576$; the
confidence is $95\,\%$, so $\alpha = 0.05$ and the lower chi-square
quantile at $29$ degrees of freedom is $\chi^{2}_{0.05, 29} = 17.71$.
Howe's factor is $k \approx 2.576\sqrt{29 \cdot (1 + 1/30)/17.71} =
2.576\sqrt{29 \cdot 1.0333/17.71} = 2.576\sqrt{1.692} = 2.576 \cdot
1.301 = 3.35$, so the tolerance interval is $12.00 \pm 3.35 \cdot 0.040
= 12.00 \pm 0.134 = [11.87, 12.13]\,\si{\milli\meter}$. The contrast
with the mean's confidence interval is the whole lesson: the $95\,\%$
confidence interval for the mean is $12.00 \pm 2.045 \cdot 0.040/
\sqrt{30} = 12.00 \pm 0.015\,\si{\milli\meter}$, nearly nine times
narrower, and an engineer who quoted it against a population-coverage
specification would certify a band that in truth holds far less than
ninety-nine percent of the pins. The tolerance factor $3.35$ against
the naive $z_{0.995} = 2.576$ is the premium the finite sample pays for
not knowing the true mean and standard deviation: were $\mu$ and
$\sigma$ known exactly, the interval would be $\mu \pm 2.576\,\sigma$,
and the extra $0.77$ in the multiplier is the estimation hedge that
vanishes only as $n \to \infty$. Figure
\ref{fig:vol02ch14:tolerance-interval} draws the three half-widths
together against sample size and shows the confidence curve diving to
zero while the prediction and tolerance curves level off above it.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-tolerance-interval.tex}

\begin{estimation}
\textbf{Question.} A pilot of $n = 20$ flexural strength tests on a
new ceramic gives sample mean $\bar{x} = 480 \,\si{\mega\pascal}$
and sample standard deviation $s = 30 \,\si{\mega\pascal}$. The
working population is modelled as Gaussian. Without consulting a
table, estimate the half-width of the $95\,\%$ confidence interval
for the population mean and the half-width of the $95\,\%$
prediction interval for the next test.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The standard error of the mean is $s/\sqrt{n} = 30/\sqrt{20}
\approx 6.7 \,\si{\mega\pascal}$. The $t$ critical value at $19$
degrees of freedom is close to $2.09$, so the confidence
half-width is about $2.09 \times 6.7 \approx 14 \,\si{\mega\pascal}$.
The prediction half-width uses $s\sqrt{1 + 1/n} \approx s = 30
\,\si{\mega\pascal}$ as the inner factor, so it is approximately
$2.09 \times 30 \approx 63 \,\si{\mega\pascal}$. The two intervals
differ by roughly the factor $\sqrt{n}$, which is the working
mnemonic.
\end{estimation}

\section{Hypothesis testing as decision theory}\pathtag{core}

Volume I Chapter 8 introduced hypothesis testing as a decision rule
under the null hypothesis. The Neyman-Pearson formulation places it
inside decision theory, where the engineer chooses a rule that
controls the probability of each kind of error and reports both
\cite[ch.~10]{text:wasserman2004}.

\subsection{The two errors, formally}

A test is a function $\delta: \mathcal{X} \to \set*{0, 1}$ of the
data, with $\delta = 1$ meaning ``reject $H_{0}$.'' The
\engterm{Type I error} probability is the rejection rate when
$H_{0}$ holds,
\[
\alpha(\delta) = \Pr_{H_{0}}[\delta(X) = 1].
\]
The \engterm{Type II error} probability is the non-rejection rate
when an alternative $H_{1}: \theta = \theta_{1}$ holds,
\[
\beta(\delta; \theta_{1}) = \Pr_{\theta_{1}}[\delta(X) = 0].
\]
The \engterm{power} of the test against $\theta_{1}$ is $1 -
\beta(\delta; \theta_{1})$. A test with low Type I error and high
power is what the engineer wants; the two cannot be made arbitrarily
small at fixed sample size, and the trade-off is governed by the
Neyman-Pearson lemma.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-test-regions.tex}

\subsection{The Neyman-Pearson lemma}

For testing a simple null $H_{0}: \theta = \theta_{0}$ against a
simple alternative $H_{1}: \theta = \theta_{1}$, the
\engterm{likelihood-ratio test}
\[
\delta^{\star}(x) = \begin{cases}
1 & \text{if } \Lambda(x) = \dfrac{p(x; \theta_{1})}{p(x; \theta_{0})} > k, \\
0 & \text{otherwise,}
\end{cases}
\]
with the threshold $k$ chosen so that $\Pr_{\theta_{0}}[\Lambda(X) >
k] = \alpha$, has the highest power among all tests of size at most
$\alpha$ \cite[ch.~10]{text:wasserman2004}. The lemma is a clean
piece of mathematics with a working consequence: under simple
hypotheses, the likelihood ratio is the optimal test statistic, and
every classical test the engineer uses is either a likelihood-ratio
test, a generalised version for composite hypotheses, or an
asymptotic approximation of one.

\subsection{Composite hypotheses, $p$-values, and the size of a test}

Real engineering tests rarely have simple alternatives. The
\engterm{generalised likelihood-ratio test} extends the lemma to
composite hypotheses by replacing $p(x; \theta_{1})$ with the
maximum of the likelihood over the alternative parameter set. The
test statistic
\[
\Lambda(x) = \frac{\sup_{\theta \in \Theta_{0}} p(x; \theta)}
                  {\sup_{\theta \in \Theta} p(x; \theta)}
\]
satisfies, under regularity conditions and at large $n$, the
asymptotic identity $-2\ln\Lambda(X) \xrightarrow{d}
\chi^{2}_{r}$, where $r$ is the difference in dimensions between
$\Theta$ and $\Theta_{0}$ \cite[ch.~10]{text:wasserman2004}. This
is the chi-squared approximation that powers most non-Gaussian
testing in working software.

The \engterm{$p$-value} of a realised observation $x$ is
\[
p(x) = \sup_{\theta \in \Theta_{0}} \Pr_{\theta}[T(X) \geq T(x)],
\]
where $T$ is the test statistic. Volume I Chapter 8's narrow
question (``how often would the null model produce a statistic this
extreme?'') is exactly the right reading; the formal definition
adds the supremum over a composite null and the explicit choice of
test statistic.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-pvalue.tex}

\subsection{Power, sample size, and design}

The dual to a fixed-$\alpha$ test is a fixed-power design. For a
one-sample $z$-test of $H_{0}: \mu = \mu_{0}$ against $H_{1}: \mu =
\mu_{0} + d\sigma$, the power at sample size $n$ is approximately
\[
\mathrm{power}(d, n; \alpha) \approx
\Phi\!\left(d\sqrt{n} - z_{1-\alpha/2}\right),
\]
where $\Phi$ is the standard normal cdf and $d$ is the
standardised effect size. Solving for $n$ at a target power
$1-\beta$ gives
\[
n \approx \left(\frac{z_{1-\alpha/2} + z_{1-\beta}}{d}\right)^{2}.
\]
A study at $\alpha = 0.05$, $1-\beta = 0.80$ needs
$n \approx (1.96 + 0.84)^{2}/d^{2} \approx 7.85/d^{2}$, the
working back-of-envelope that detecting a half-sigma effect
demands about $30$ observations and detecting a fifth-sigma effect
demands about $200$. The power curves in figure
\ref{fig:vol02:ch14:power-curve} make the trade explicit.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-power-curve.tex}

\paragraph{Worked example: sizing a two-proportion reliability
trial.} A redesign is meant to lower a component's field-failure
rate from a baseline $p_{1} = 0.10$ to a target $p_{2} = 0.06$, and
the engineer must size the comparison trial so the improvement, if
real, is detected with power $0.80$ at $\alpha = 0.05$ two-sided.
For two independent proportions the per-group sample size is
\[
n \approx \frac{\bigl(z_{1-\alpha/2}\sqrt{2\bar{p}(1-\bar{p})}
+ z_{1-\beta}\sqrt{p_{1}(1-p_{1}) + p_{2}(1-p_{2})}\bigr)^{2}}
{(p_{1} - p_{2})^{2}},
\]
with $\bar{p} = (p_{1}+p_{2})/2 = 0.08$. The pooled term is
$\sqrt{2 \cdot 0.08 \cdot 0.92} = \sqrt{0.1472} = 0.384$ and the
unpooled term is $\sqrt{0.10 \cdot 0.90 + 0.06 \cdot 0.94} =
\sqrt{0.090 + 0.0564} = \sqrt{0.1464} = 0.383$. The numerator is
$(1.96 \cdot 0.384 + 0.84 \cdot 0.383)^{2} = (0.753 + 0.322)^{2}
= 1.075^{2} = 1.156$, and the denominator is $(0.04)^{2} = 0.0016$,
so $n \approx 1.156/0.0016 = 722$ components per group, about
$1450$ in all. The number is large because the effect is a
four-percentage-point shift on small rates, and small-probability
differences carry little Fisher information per unit. An engineer
who proposed a fifty-unit trial for this comparison would be
running a study with power near $0.10$, almost certain to miss a
real improvement, which is the design failure the calculation
exists to prevent before the trial is run, not after it returns a
non-significant result.

A study that reports a non-significant result and reports its
power against the effect size the engineer cares about is an
honest report. A study that reports a non-significant result and
does not report power is reporting that no effect was detected
and is silent on whether the design could have detected one.

\subsection{The classical tests as worked instances}

Three tests cover most of the engineer's day-to-day inference, and
each is a generalised likelihood-ratio test or a close
approximation of one. We work each on numbers.

\paragraph{One-sample and two-sample $t$-tests.} For a single
sample with unknown variance, the statistic for $H_{0}: \mu =
\mu_{0}$ is $T = (\bar{X} - \mu_{0})/(S/\sqrt{n})$, which follows
$t_{n-1}$ under the null. Suppose a batch of $n = 16$ machined
shafts has target diameter $\mu_{0} = 25.00\,\si{\milli\meter}$,
measured mean $\bar{x} = 25.06\,\si{\milli\meter}$, and sample
standard deviation $s = 0.10\,\si{\milli\meter}$. The statistic is
$T = (25.06 - 25.00)/(0.10/4) = 0.06/0.025 = 2.40$. Against the
two-sided critical value $t_{0.975, 15} = 2.131$, the test rejects:
the batch mean is inconsistent with the target at $\alpha = 0.05$.
The two-sided $p$-value is $\Pr[\abs{t_{15}} > 2.40] \approx
0.030$. The engineer reports the rejection together with the effect
size $(\bar{x} - \mu_{0})/s = 0.6$ standard deviations, which is the
quantity the downstream tolerance decision actually needs.

For two independent samples with a common variance, the
\engterm{pooled-variance two-sample $t$-test} uses $S_{p}^{2} =
\frac{(n_{1}-1)S_{1}^{2} + (n_{2}-1)S_{2}^{2}}{n_{1} + n_{2} - 2}$
and statistic $T = (\bar{X}_{1} - \bar{X}_{2})/(S_{p}\sqrt{1/n_{1}
+ 1/n_{2}})$ on $n_{1} + n_{2} - 2$ degrees of freedom. With
$\bar{x}_{1} = 480$, $s_{1} = 30$, $n_{1} = 20$ and $\bar{x}_{2} =
465$, $s_{2} = 25$, $n_{2} = 18$ (two ceramic batches), the pooled
variance is $\frac{19 \cdot 900 + 17 \cdot 625}{36} = \frac{17100
+ 10625}{36} = 770.1$, so $s_{p} = 27.75$. The statistic is
$T = 15/(27.75\sqrt{1/20 + 1/18}) = 15/(27.75 \cdot 0.3249) =
15/9.02 = 1.66$, below $t_{0.975, 36} \approx 2.028$. The two-sided
$p$-value is about $0.105$, so the batches are not distinguishable
at $\alpha = 0.05$; the engineer reports a difference of $15\,
\si{\mega\pascal}$ with a $95\,\%$ confidence interval of $15 \pm
2.028 \cdot 9.02 = 15 \pm 18.3\,\si{\mega\pascal}$, which contains
zero. The interval, not the binary verdict, is the deliverable.

\paragraph{The chi-square goodness-of-fit and independence tests.}
For a categorical variable with $k$ cells, observed counts $O_{j}$
and expected counts $E_{j}$ under the null, the
\engterm{Pearson chi-square statistic} is $\chi^{2} = \sum_{j=1}^{k}
(O_{j} - E_{j})^{2}/E_{j}$, asymptotically $\chi^{2}_{k-1-m}$ where
$m$ is the number of parameters estimated from the data. A die
rolled $n = 120$ times yields counts $\set{18, 24, 16, 22, 20, 20}$
against the fair-die expectation $E_{j} = 20$ for each face. The
statistic is $[(18-20)^{2} + (24-20)^{2} + (16-20)^{2} + (22-20)^{2}
+ 0 + 0]/20 = (4 + 16 + 16 + 4)/20 = 40/20 = 2.00$ on $k - 1 = 5$
degrees of freedom. Against $\chi^{2}_{0.95, 5} = 11.07$, the test
does not reject: the data are consistent with a fair die, with
$p \approx 0.85$. For a two-way contingency table testing
independence, the expected count in cell $(i, j)$ is
$E_{ij} = R_{i}C_{j}/n$ (row total times column total over the
grand total), and the degrees of freedom are $(r-1)(c-1)$. The
chi-square statistic is the generalised likelihood-ratio statistic's
quadratic approximation, which is why the two agree to leading order
on large tables.

\paragraph{Worked example: an independence test on a $2\times3$ table.}
A quality engineer asks whether the defect type on a rejected part is
associated with the shift that produced it. The counts across three
defect types and two shifts are

\begin{center}
\small
\begin{tabular}{lcccc}
\toprule
        & crack & void & inclusion & row total \\
\midrule
day shift   & $30$ & $20$ & $10$ & $60$ \\
night shift & $18$ & $22$ & $20$ & $60$ \\
\midrule
column total & $48$ & $42$ & $30$ & $120$ \\
\bottomrule
\end{tabular}
\end{center}

Under the null of independence the expected count in each cell is the
row total times the column total over the grand total, $E_{ij} =
R_{i}C_{j}/n$. With equal row totals of $60$ each, the expected counts
are the column totals halved: $E_{\cdot,\text{crack}} = 24$,
$E_{\cdot,\text{void}} = 21$, $E_{\cdot,\text{inclusion}} = 15$ in both
rows. The Pearson statistic sums $(O - E)^{2}/E$ over the six cells:
day shift contributes $(30-24)^{2}/24 + (20-21)^{2}/21 +
(10-15)^{2}/15 = 1.500 + 0.048 + 1.667 = 3.215$, and night shift
contributes the same magnitudes by the symmetry of the equal row
totals, $(18-24)^{2}/24 + (22-21)^{2}/21 + (20-15)^{2}/15 = 1.500 +
0.048 + 1.667 = 3.215$, for a total $\chi^{2} = 6.43$ on $(r-1)(c-1) =
1 \cdot 2 = 2$ degrees of freedom. Against $\chi^{2}_{0.95,2} = 5.99$
the test rejects at $\alpha = 0.05$ with $p \approx 0.040$: defect type
and shift are associated, and the table shows the direction, night
shift running heavier on inclusions and lighter on cracks. The engineer
reports the association together with the largest cell contributions,
because the two inclusion cells and the two crack cells carry almost
all of the statistic and point the follow-up investigation at the
night-shift inclusion mechanism rather than at voids, whose
contribution is negligible.

\paragraph{One-way analysis of variance.} When the engineer
compares $g$ group means, running $\binom{g}{2}$ pairwise
$t$-tests inflates the family-wise error rate, the very failure
section 14.7 catalogues. \engterm{One-way ANOVA} tests the single
null $H_{0}: \mu_{1} = \cdots = \mu_{g}$ by decomposing the total
sum of squares into a between-group part and a within-group part:
\[
\underbrace{\sum_{i,j}(Y_{ij} - \bar{Y})^{2}}_{\mathrm{SS_{tot}}}
= \underbrace{\sum_{i} n_{i}(\bar{Y}_{i} - \bar{Y})^{2}}_{\mathrm{SS_{between}}}
+ \underbrace{\sum_{i,j}(Y_{ij} - \bar{Y}_{i})^{2}}_{\mathrm{SS_{within}}}.
\]
The test statistic is the ratio of mean squares, $F = \mathrm{MS_{between}}
/ \mathrm{MS_{within}}$, where $\mathrm{MS_{between}} =
\mathrm{SS_{between}}/(g-1)$ and $\mathrm{MS_{within}} =
\mathrm{SS_{within}}/(N - g)$, distributed as $F_{g-1, N-g}$ under
the null. Three catalysts tested at $\bar{y}_{1} = 72$,
$\bar{y}_{2} = 78$, $\bar{y}_{3} = 75$ on $n_{i} = 10$ each
($N = 30$, grand mean $\bar{y} = 75$) give $\mathrm{SS_{between}} =
10[(72-75)^{2} + (78-75)^{2} + 0] = 10(9 + 9) = 180$,
$\mathrm{MS_{between}} = 180/2 = 90$. With a within-group variance
of $s^{2} = 20$ pooled across groups, $\mathrm{MS_{within}} = 20$
and $\mathrm{SS_{within}} = 20 \cdot 27 = 540$, so $F = 90/20 =
4.50$ on $(2, 27)$ degrees of freedom. Against $F_{0.95, 2, 27}
= 3.35$, the test rejects: at least one catalyst mean differs. A
significant ANOVA licenses a follow-up multiple-comparisons
procedure (Tukey's honest significant difference, treated in
Volume I Chapter 8) to identify which pairs differ, with the
family-wise error rate controlled.

\paragraph{Two-way analysis of variance and the blocked design.}
The one-way decomposition extends to two crossed factors, and the
extension is where analysis of variance earns its place in
experimental design \cite[ch.~5--7]{text:box-hunter-hunter-2005}.
With factor A at $a$ levels, factor B at $b$ levels, and $r$
replicates per cell, the model partitions each observation
$Y_{ijk}$ as a grand mean, a row effect, a column effect, an
interaction, and noise:
\[
Y_{ijk} = \mu + \alpha_{i} + \beta_{j} + (\alpha\beta)_{ij}
        + \varepsilon_{ijk},
\qquad \varepsilon_{ijk} \sim \mathcal{N}(0, \sigma^{2}).
\]
The total sum of squares decomposes into four orthogonal parts,
$\mathrm{SS_{tot}} = \mathrm{SS_{A}} + \mathrm{SS_{B}} +
\mathrm{SS_{AB}} + \mathrm{SS_{E}}$, and each factor gets its own
$F$-ratio against the error mean square. The interaction term is the
piece a one-way analysis cannot see: it measures whether the effect
of factor A depends on the level of factor B.

Consider a fatigue experiment crossing three alloys (factor A) with
two coatings (factor B), $r = 3$ specimens per cell, response the
fatigue life in kilocycles. The cell means are

\begin{center}
\small
\begin{tabular}{lccc}
\toprule
        & A1 & A2 & A3 \\
\midrule
coating B1 & $52$ & $58$ & $66$ \\
coating B2 & $48$ & $60$ & $74$ \\
\bottomrule
\end{tabular}
\end{center}

with a within-cell error mean square $\mathrm{MS_{E}} = 12.0$ on
$ab(r-1) = 3 \cdot 2 \cdot 2 = 12$ degrees of freedom. The grand
mean is $\bar{Y} = 59.67$. The row (coating) means are
$\bar{Y}_{B1} = 58.67$ and $\bar{Y}_{B2} = 60.67$; the column
(alloy) means are $\bar{Y}_{A1} = 50$, $\bar{Y}_{A2} = 59$,
$\bar{Y}_{A3} = 70$. The alloy sum of squares is $\mathrm{SS_{A}}
= br\sum_{i}(\bar{Y}_{Ai} - \bar{Y})^{2} = 2 \cdot 3\,
[(50 - 59.67)^{2} + (59 - 59.67)^{2} + (70 - 59.67)^{2}]
= 6\,(93.4 + 0.45 + 106.7) = 6 \cdot 200.6 = 1203.3$ on
$a - 1 = 2$ degrees of freedom, so $\mathrm{MS_{A}} = 601.7$ and
$F_{A} = 601.7/12.0 = 50.1$, far above $F_{0.95,2,12} = 3.89$:
alloy matters. The coating sum of squares is $\mathrm{SS_{B}} =
ar\sum_{j}(\bar{Y}_{Bj} - \bar{Y})^{2} = 3 \cdot 3\,[(58.67 -
59.67)^{2} + (60.67 - 59.67)^{2}] = 9 \cdot 2.0 = 18.0$ on $1$
degree of freedom, so $F_{B} = 18.0/12.0 = 1.5$, below
$F_{0.95,1,12} = 4.75$: the coating main effect is not significant.
The interaction sum of squares, $\mathrm{SS_{AB}} = r\sum_{i,j}
(\bar{Y}_{ij} - \bar{Y}_{Ai} - \bar{Y}_{Bj} + \bar{Y})^{2}$,
evaluates to $48.0$ on $(a-1)(b-1) = 2$ degrees of freedom, so
$\mathrm{MS_{AB}} = 24.0$ and $F_{AB} = 24.0/12.0 = 2.0$, below
$F_{0.95,2,12} = 3.89$ but large enough to warrant the
interaction plot. The lesson the table carries is the order of
reading: a significant interaction is interpreted before the main
effects, because a coating effect averaged over alloys (the
non-significant $F_{B}$) hides that coating B2 adds $8$ kilocycles
at A3 and subtracts $4$ at A1. Figure
\ref{fig:vol02:ch14:twoway-interaction} draws the cell means and
shows the fan that the averaged main effect conceals.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-twoway-interaction.tex}

The \engterm{randomised block design} is the two-way analysis put
to work against nuisance variation. When specimens come from
several heats of steel, several test machines, or several days, the
block (heat, machine, day) is entered as a second factor whose
effect the engineer does not care about but must remove. Blocking
moves the block-to-block variation out of the error term and into
its own sum of squares, shrinking $\mathrm{MS_{E}}$ and raising the
power of the test on the factor of interest. The working rule from
the design literature is direct: block what you can, randomise what
you cannot \cite[ch.~4]{text:box-hunter-hunter-2005}. A treatment
comparison run without blocking against a known nuisance variable
pays for the omission in inflated error variance and lost power.

\paragraph{The permutation test.} When the Gaussian assumption
behind the $t$-test is in doubt and the sample is too small to lean
on the central limit theorem, the \engterm{permutation test}
supplies a $p$-value that assumes only \engterm{exchangeability}:
under the null, the group labels carry no information, so any
relabelling of the pooled data is as likely as the observed one.
The procedure pools the two groups, recomputes the test statistic
(say the difference in group means) under many random relabellings,
and reports the fraction of relabelled statistics at least as
extreme as the observed one. For two groups of sizes $n_{A}$ and
$n_{B}$ the exact test enumerates all $\binom{n_{A}+n_{B}}{n_{A}}$
splits; for larger samples a Monte Carlo approximation with $B$
random relabellings suffices, with the reported $p$-value carrying
its own Monte Carlo standard error $\sqrt{p(1-p)/B}$.

Suppose group A has measurements $\set{12, 15, 14, 17, 13}$ (mean
$14.2$) and group B has $\set{9, 11, 8, 10}$ (mean $9.5$), a raw
difference of $4.7$ working units. The pooled nine values are
relabelled into a group of five and a group of four; for each of
$B = 2000$ relabellings the difference in means is recomputed, and
the observed $4.7$ is compared against the resulting reference
distribution. Figure \ref{fig:vol02:ch14:permutation-null} shows the
reference distribution centred on zero with the observed value far
in the upper tail; for data this separated the two-sided
permutation $p$-value lands near $0.01$ to $0.02$, in agreement with
the Welch $t$-test but obtained without the normality assumption.
The permutation test is the engineer's default check when a
$t$-test result sits near the decision threshold and the sample is
small enough that the tail behaviour of the $t$ approximation is
itself uncertain.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-permutation-null.tex}

\begin{mastery}
The permutation test and the bootstrap answer different questions
with the same resampling machinery, and conflating them is a common
error. The permutation test resamples \emph{without} replacement by
relabelling, and tests a sharp null of exchangeability (the two
distributions are identical); its $p$-value is exact under that null
regardless of sample size. The bootstrap resamples \emph{with}
replacement from each group separately, and estimates the sampling
distribution of a statistic (a standard error, a confidence
interval) without committing to any null. The permutation test is
the right tool when the question is ``could the label assignment be
chance?''; the bootstrap is the right tool when the question is ``how
uncertain is this estimate?'' A permutation confidence interval and a
bootstrap hypothesis test both exist but are the less natural
direction for each machine. The shared limitation is that both lean
on the data carrying the relevant structure: a permutation test on
paired data must permute within pairs, and a bootstrap on dependent
data must resample blocks, or both procedures silently assume an
independence the data do not have.
\end{mastery}

\subsection{Equivalence testing: proving two things are the same}

A significance test that fails to reject its null has not shown the
two groups are equal; it has shown the data do not rule equality out,
which is a different and weaker statement. The engineer who must
demonstrate that a cheaper supplier's material is \emph{as good as}
the incumbent, or that a redesign did not degrade a response, needs a
test whose rejection means sameness, not a test whose non-rejection is
read, wrongly, as sameness. \engterm{Equivalence testing} supplies it
by inverting the burden of proof. The engineer states an
\engterm{equivalence margin} $\Delta$, the largest difference that is
practically negligible, and tests the compound null that the true
difference lies \emph{outside} $[-\Delta, +\Delta]$ against the
alternative that it lies inside. Rejecting that null establishes
equivalence within the margin, at the stated level.

The standard machine is the \engterm{two one-sided tests} procedure,
or TOST. The interval null $\abs{\mu_{1} - \mu_{2}} \geq \Delta$ splits
into two one-sided pieces, $\mu_{1} - \mu_{2} \leq -\Delta$ and
$\mu_{1} - \mu_{2} \geq +\Delta$, and each is tested at level $\alpha$
with a one-sided $t$-test. Equivalence is declared only when both
one-sided tests reject, which happens exactly when the ordinary
$(1 - 2\alpha)$ confidence interval for the difference falls entirely
inside $[-\Delta, +\Delta]$. The $1 - 2\alpha$ rather than $1 - \alpha$
level is not a slip: the two one-sided tests at $\alpha$ each control
their own tail, and their conjunction controls the equivalence claim
at $\alpha$ while corresponding to a $90\,\%$ interval when $\alpha =
0.05$.

\paragraph{Worked example: qualifying a substitute alloy.} A qualified
alloy has a mean tensile strength the engineer will accept a shortfall
of no more than $\Delta = 10\,\si{\mega\pascal}$ against; a candidate
substitute must be shown equivalent within that margin. Twenty
specimens of each give $\bar{x}_{\text{inc}} = 465$, $\bar{x}_{\text{cand}}
= 461$, a difference of $4\,\si{\mega\pascal}$, with a pooled standard
error of the difference of $\widehat{\mathrm{se}} = 2.6\,
\si{\mega\pascal}$ on $38$ degrees of freedom. The naive reading notes
that an ordinary two-sided $t$-test of $H_{0}: \mu_{\text{inc}} =
\mu_{\text{cand}}$ gives $t = 4/2.6 = 1.54$, $p \approx 0.13$, fails to
reject, and stops, concluding nothing. The equivalence test does the
work the naive test cannot. The upper one-sided statistic for
$H_{0}: \mu_{\text{inc}} - \mu_{\text{cand}} \geq 10$ is $t_{U} = (4 -
10)/2.6 = -2.31$, which rejects at $\alpha = 0.05$ against the lower
critical value $-t_{0.95, 38} = -1.686$; the lower one-sided statistic
for $H_{0}: \mu_{\text{inc}} - \mu_{\text{cand}} \leq -10$ is $t_{L} =
(4 + 10)/2.6 = 5.38$, which rejects easily. Both one-sided tests
reject, so equivalence within $10\,\si{\mega\pascal}$ is established at
$\alpha = 0.05$. Equivalently, the $90\,\%$ confidence interval for the
difference is $4 \pm 1.686 \cdot 2.6 = [-0.38, 8.38]\,\si{\mega\pascal}$,
which sits inside $[-10, 10]$, and the substitute qualifies. The
lesson is the asymmetry the two frames expose: the ordinary test's
$p = 0.13$ was silent on equivalence, and an engineer who had read
its non-rejection as ``the alloys are the same'' would have made a
claim the data did not support, in the same direction and for the same
reason that a non-significant safety test is silent on whether a
hazard is absent.

\subsection{Sequential testing: deciding as the data arrive}

The tests so far fix the sample size in advance and decide once. When
observations arrive one at a time and each new unit is costly, the
engineer can often decide sooner by testing after every observation,
provided the stopping rule is designed to hold the error rates. Naive
repeated testing inflates the false-alarm rate exactly as
multiple comparisons do, so the sequential procedure must be built to
account for the repeated looks. The \engterm{sequential probability
ratio test}, or SPRT, does this for a simple null against a simple
alternative by accumulating the log-likelihood ratio and comparing it
to two boundaries \cite[ch.~10]{text:wasserman2004}. After $m$
observations the running statistic is
\[
\Lambda_{m} = \sum_{i=1}^{m} \ln\frac{p(x_{i}; \theta_{1})}
{p(x_{i}; \theta_{0})},
\]
and the rule accepts $H_{0}$ when $\Lambda_{m} \leq \ln B$, accepts
$H_{1}$ when $\Lambda_{m} \geq \ln A$, and takes one more observation
otherwise. The boundaries follow from the target error rates by Wald's
approximation, $A \approx (1-\beta)/\alpha$ and $B \approx \beta/
(1-\alpha)$, and the test's payoff is that its expected sample size
under either hypothesis is typically well below the fixed-sample size
that would achieve the same $\alpha$ and $\beta$.

\paragraph{Worked example: an accept-on-quality sampling test.} An
incoming lot is acceptable at a defect rate $p_{0} = 0.02$ and
rejectable at $p_{1} = 0.10$, and the inspector wants $\alpha = 0.05$,
$\beta = 0.10$. The boundaries are $A = 0.90/0.05 = 18$ and
$B = 0.10/0.95 = 0.105$, so $\ln A = 2.89$ and $\ln B = -2.25$. Each
inspected unit contributes to the running log-likelihood ratio $\ln
(p_{1}/p_{0}) = \ln(0.10/0.02) = 1.609$ if it is defective and
$\ln((1 - p_{1})/(1 - p_{0})) = \ln(0.90/0.98) = -0.0851$ if it is
sound. Inspecting a stream that runs clean, each good unit subtracts
$0.0851$ from the statistic, and after $\lceil 2.25/0.0851 \rceil = 27$
consecutive good units the statistic crosses $\ln B$ and the lot is
accepted, well before a fixed-sample plan of comparable power would
finish. A single early defect adds $1.609$ and pushes the decision the
other way: two defects in the first handful of units drive the
statistic toward $\ln A = 2.89$ and reject the lot on a tiny sample.
The expected number inspected under a clean lot is near $30$ against
a fixed-sample plan needing on the order of $80$, and the saving is the
reason acceptance sampling adopted the sequential design where units
are expensive to test. The discipline the SPRT enforces is that the
two boundaries, not a fixed count, define the stopping rule, and they
were fixed from $\alpha$ and $\beta$ before the first unit was
inspected, which is the same pre-specification discipline the failure
section of this chapter demands of every analysis.

\subsection{Decision-theoretic framing}

The engineer who frames a test as a decision under cost is closer to
practice than the engineer who reports $p$-values alone. Let $C_{I}$
be the cost of a Type I error (rejecting a true null) and $C_{II}$
the cost of a Type II error. Under a prior probability $\pi_{0}$ on
$H_{0}$, the expected cost of test $\delta$ is
\[
\E[C(\delta)] = \pi_{0}\, C_{I}\, \alpha(\delta)
              + (1 - \pi_{0})\, C_{II}\, \beta(\delta).
\]
Choosing $\delta$ to minimise expected cost gives the Bayes-optimal
test. The likelihood-ratio threshold then has a closed form,
$k = (\pi_{0} C_{I})/((1 - \pi_{0}) C_{II})$, which makes the
$\alpha$-versus-power trade explicit and quantitative. A
safety-critical decision is not ``smaller $\alpha$'' by reflex; it
is the test whose threshold reflects the actual cost ratio.

\subsection{Applied vignette: statistical process control as a live test}

The decision-theoretic framing becomes a running procedure the moment
the engineer stops testing a single sample and starts testing every
sample as it arrives. A fill line dispenses a nominal $100\,\si{\gram}$
of product into each container, and the operator wants to know, in real
time, whether the process mean has drifted off target. The tool is the
\engterm{Shewhart control chart}, and reading it as a sequence of
hypothesis tests is what connects it to this section.

The process is sampled in \engterm{rational subgroups}: every fifteen
minutes the operator weighs $n = 5$ consecutive containers and records
the subgroup mean $\bar{x}$. Under the in-control null, the individual
fills are $\mathcal{N}(100, \sigma^{2})$ with a process standard
deviation established from a capability study as $\sigma = 4.47\,
\si{\gram}$, so the subgroup mean has standard error
$\sigma/\sqrt{n} = 4.47/\sqrt{5} = 2.00\,\si{\gram}$. The
\engterm{control limits} sit at three standard errors either side of
the target, $100 \pm 3 \cdot 2.00 = [94, 106]\,\si{\gram}$. Each
plotted subgroup mean is then a two-sided test of $H_{0}: \mu = 100$
at a per-point significance level of $\Pr[\abs{Z} > 3] = 0.0027$, the
false-alarm rate the three-sigma convention encodes. The choice of
three sigma rather than two is a decision-theoretic one: a fill line
that stops on every alarm pays for a tighter limit in downtime, so the
limit is set where the cost of a false stop and the cost of a missed
shift balance, and three sigma is the working compromise the process
literature settled on \cite[ch.~4]{text:box-hunter-hunter-2005}.

Figure \ref{fig:vol02:ch14:control-chart} plots twenty subgroup means.
The first nine sit inside the limits, consistent with the null. The
tenth pierces the upper limit at $105.8\,\si{\gram}$, a single-point
out-of-control signal that triggers an immediate investigation: at a
per-point false-alarm rate of $0.0027$, a point outside the limits is
far more plausibly a real shift than a chance excursion, and the
operator acts on it. The subtler signal is the run of eight rising
points from subgroup thirteen onward, none of which leaves the band.
No single point rejects its null, yet the probability that eight
consecutive independent points all fall above the centre line under the
in-control null is $(1/2)^{8} = 0.0039$, comparable to the single-point
alarm rate, so the \engterm{run rules} that supplement the three-sigma
limit flag the drift before any point escapes. The chart catches a slow
mean shift that a sequence of independent single-point tests, each
looking only at its own subgroup, would miss.

The economic reading closes the vignette. The \engterm{average run
length} is the expected number of subgroups until an alarm. In control,
it is $1/0.0027 \approx 370$ subgroups between false alarms, roughly
one false stop per four days at fifteen-minute sampling, a rate the
line can absorb. When the mean shifts by one standard error, to
$102\,\si{\gram}$, the probability a subgroup mean exceeds the upper
limit rises to $\Pr[Z > (106 - 102)/2] = \Pr[Z > 2] = 0.023$, giving
an out-of-control average run length near $1/0.023 \approx 44$
subgroups, about eleven hours to detect a one-sigma shift. An engineer
who needs faster detection of small shifts reaches for the cumulative
sum or exponentially weighted moving average charts, which pool
information across subgroups and detect a one-sigma shift in a handful
of points rather than dozens, at the cost of a delayed response to a
large sudden shift. The choice among the charts is the same
decision-theoretic trade this section opened with: the cost of a false
alarm against the cost of a slow one, made explicit and quantified
rather than defaulted.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-control-chart.tex}

\section{Likelihood and maximum likelihood}\pathtag{core}

Given data $X_{1}, \ldots, X_{n}$ with joint density $p(x_{1},
\ldots, x_{n}; \theta)$, the \engterm{likelihood function} treats
the data as fixed and the parameter as the variable:
\[
L(\theta) = p(x_{1}, \ldots, x_{n}; \theta).
\]
For independent observations, the likelihood factorises:
$L(\theta) = \prod_{i=1}^{n} p(x_{i}; \theta)$. The
\engterm{log-likelihood} $\ell(\theta) = \log L(\theta)$ converts
the product into a sum and is the working object in nearly every
calculation. The \engterm{maximum-likelihood estimator}, or MLE, is
\[
\hat{\theta}_{\mathrm{MLE}} = \argmax_{\theta \in \Theta} \ell(\theta).
\]

\subsection{MLE for canonical distributions}

The MLE is computable in closed form for several distributions. We
work the cases the engineer meets most often; Devore's standard
applied-statistics catalogue covers the same families with
worked-engineering data \cite[ch.~6, ch.~7]{text:devore2015}.

\textit{Bernoulli}. For $X_{i} \in \set*{0, 1}$ with $\Pr[X = 1] =
p$, the log-likelihood is $\ell(p) = (\sum x_{i})\log p +
(n - \sum x_{i})\log(1-p)$. Setting $\dd\ell/\dd p = 0$ gives
$\hat{p} = \bar{x}$. The MLE recovers the sample proportion.

\textit{Gaussian with known $\sigma$}. For $X_{i} \sim
\mathcal{N}(\mu, \sigma^{2})$ with $\sigma$ known, the
log-likelihood is $\ell(\mu) = -\frac{1}{2\sigma^{2}}\sum
(x_{i}-\mu)^{2} + \text{const}$. The MLE for $\mu$ is the sample
mean $\bar{x}$.

\textit{Gaussian with both unknown}. The MLE for the variance is
$\hat{\sigma}^{2}_{\mathrm{MLE}} = \frac{1}{n}\sum(x_{i} -
\bar{x})^{2}$, with denominator $n$, not $n-1$. The MLE is biased
for $\sigma^{2}$; the Bessel-corrected estimator is unbiased. The
two converge as $n$ grows.

\textit{Exponential}. For $X_{i} \sim \mathrm{Exp}(\lambda)$, the
log-likelihood is $\ell(\lambda) = n\log\lambda - \lambda \sum
x_{i}$. The MLE is $\hat{\lambda} = 1/\bar{x}$.

\textit{Poisson}. For $X_{i} \sim \mathrm{Poisson}(\lambda)$, the
MLE is $\hat{\lambda} = \bar{x}$.

\textit{Uniform on $[0, \theta]$}. For $X_{i} \sim
\mathrm{Uniform}[0, \theta]$, the likelihood is zero for $\theta <
\max x_{i}$ and decreasing in $\theta$ for $\theta \geq \max x_{i}$.
The MLE is $\hat{\theta} = \max x_{i}$. The example shows that the
MLE need not be a sum of simple terms: the support of the
distribution carries information that the score equation does not.

\paragraph{Worked derivation: Gaussian with both parameters unknown.}
The full two-parameter Gaussian MLE earns a complete derivation
because it is the workhorse of regression and process control. The
log-likelihood for $X_{i} \sim \mathcal{N}(\mu, \sigma^{2})$ is
\[
\ell(\mu, \sigma^{2}) = -\frac{n}{2}\log(2\pi) - \frac{n}{2}
\log\sigma^{2} - \frac{1}{2\sigma^{2}}\sum_{i=1}^{n}(x_{i} - \mu)^{2}.
\]
The score for $\mu$ is $\pdv{\ell}{\mu} = \frac{1}{\sigma^{2}}
\sum_{i}(x_{i} - \mu)$, which vanishes at $\hat{\mu} = \bar{x}$
regardless of $\sigma^{2}$. Substituting $\hat{\mu}$ and taking
the score for $\sigma^{2}$,
\[
\pdv{\ell}{\sigma^{2}} = -\frac{n}{2\sigma^{2}} + \frac{1}{2\sigma^{4}}
\sum_{i}(x_{i} - \bar{x})^{2} = 0
\;\Longrightarrow\;
\hat{\sigma}^{2}_{\mathrm{MLE}} = \frac{1}{n}\sum_{i}(x_{i} - \bar{x})^{2}.
\]
The MLE for the mean is the sample mean and the MLE for the
variance is the uncorrected sample variance, with denominator $n$.
The Hessian at the maximum is block diagonal between $\mu$ and
$\sigma^{2}$ (the off-diagonal expectation is zero because the
score for $\mu$ has mean zero), which is why the two estimators are
asymptotically independent and the standard error of $\bar{x}$ does
not depend on the uncertainty in $\hat{\sigma}^{2}$ to leading
order. The Fisher information matrix is
$I_{1}(\mu, \sigma^{2}) = \operatorname{diag}(1/\sigma^{2},
1/(2\sigma^{4}))$, so the asymptotic variances are $\sigma^{2}/n$
for $\hat{\mu}$ and $2\sigma^{4}/n$ for $\hat{\sigma}^{2}$.

Figure \ref{fig:vol02:ch14:likelihood-surface} draws the log-likelihood
as a contour map over the two parameters, and the geometry reads the
algebra back. The vertical ridge at $\mu = \bar{x}$ is the statement
that the mean estimate does not move with the variance; the level sets
stretch further along the $\sigma$ axis than the $\mu$ axis, the
picture of a variance estimate whose sampling distribution is skewed at
small $n$; and the single interior peak is the joint MLE the score
equations located. An engineer who has met numerical optimisation
reads the same figure as a well-conditioned maximisation, concave near
the top with no competing local maxima, which is why the Gaussian MLE
is found in one step by any reasonable solver.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-likelihood-surface.tex}

\paragraph{Worked example: an exponential lifetime MLE with its
standard error.} A reliability test records ten times-to-failure
(in thousands of hours): $\set{0.7, 1.9, 0.4, 2.6, 1.1, 0.9, 3.2,
0.5, 1.4, 0.8}$, summing to $13.5$, so $\bar{x} = 1.35$. Under
$X_{i} \sim \mathrm{Exp}(\lambda)$ the MLE is $\hat{\lambda} =
1/\bar{x} = 0.741$ per thousand hours. The per-observation Fisher
information for the exponential is $I_{1}(\lambda) = 1/\lambda^{2}$,
so $I_{n}(\hat{\lambda}) = n/\hat{\lambda}^{2} = 10/0.549 = 18.2$
and the asymptotic standard error is $\hat{\lambda}/\sqrt{n} =
0.741/\sqrt{10} = 0.234$. A $95\,\%$ Wald interval for $\lambda$
is $0.741 \pm 1.96 \cdot 0.234 = [0.282, 1.200]$. The engineer who
needs the mean lifetime $1/\lambda$ should invert the endpoints
rather than the point estimate, because the Wald interval is not
transformation-invariant: the interval for the mean lifetime is
$[1/1.200, 1/0.282] = [0.83, 3.55]$ thousand hours, which is
asymmetric about the point estimate $1.35$. The asymmetry is real
and is the signature of a parameter whose sampling distribution is
skewed at small $n$.

\subsection{The score and the Fisher information}

The \engterm{score function} is the gradient of the log-likelihood,
$U(\theta) = \pdv{\ell(\theta)}{\theta}$. Under regularity, the
score has mean zero at the true parameter. The \engterm{Fisher
information} is
\[
I(\theta) = \E_{\theta}\!\left[U(\theta)^{2}\right]
          = -\E_{\theta}\!\left[\pdv{^{2}\ell(\theta)}{\theta^{2}}\right],
\]
where the second equality holds under regularity. For a sample of
size $n$ from an i.i.d.\ model, the Fisher information scales
linearly: $I_{n}(\theta) = n I_{1}(\theta)$.

\subsection{Properties of the MLE}

Under regularity conditions (the parameter set is open, the support
of the distribution does not depend on the parameter, the
log-likelihood is twice differentiable, and standard moment
conditions hold), the MLE satisfies three working properties
\cite[ch.~9]{text:wasserman2004}.

\textit{Consistency}. $\hat{\theta}_{\mathrm{MLE}} \xrightarrow{P}
\theta$ as $n \to \infty$.

\textit{Asymptotic normality}. $\sqrt{n}(\hat{\theta}_{\mathrm{MLE}}
- \theta) \xrightarrow{d} \mathcal{N}(0, I_{1}(\theta)^{-1})$. The
asymptotic variance is the inverse of the per-observation Fisher
information.

\textit{Asymptotic efficiency}. No regular asymptotically unbiased
estimator has smaller asymptotic variance than the MLE. The
\engterm{Cram\'er-Rao lower bound} states that any unbiased
estimator $\hat{\theta}$ satisfies $\Var(\hat{\theta}) \geq
1/I_{n}(\theta)$, and the MLE attains this bound asymptotically.

The three properties together justify the working habit: when the
model is right and $n$ is large, the MLE with its observed-Fisher
standard error is the engineer's first reach.

\begin{mastery}
The Cram\'er-Rao bound has a two-line proof that exposes why
information and variance are reciprocal. For an unbiased estimator
$\hat{\theta}$, differentiating $\E[\hat{\theta}] = \theta$ under
the integral gives $\E[\hat{\theta}\,U(\theta)] = 1$, where
$U = \pdv{\ell}{\theta}$ is the score with mean zero. Hence
$\Cov(\hat{\theta}, U) = 1$. The Cauchy-Schwarz inequality applied
to the covariance gives $1 = \Cov(\hat{\theta}, U)^{2} \leq
\Var(\hat{\theta})\Var(U) = \Var(\hat{\theta})\,I(\theta)$, so
$\Var(\hat{\theta}) \geq 1/I(\theta)$. Equality holds when
$\hat{\theta} - \theta$ is proportional to the score, which happens
exactly for the canonical exponential-family estimators (the sample
mean for a Gaussian mean, the sample proportion for a Bernoulli).
The bound is the formal reason the engineer cannot keep shrinking an
unbiased estimator's variance: information is finite, and the data
carry only so much of it.
\end{mastery}

\subsection{The Wald, score, and likelihood-ratio trio}

Three tests of $H_{0}: \theta = \theta_{0}$ fall out of the
likelihood, and they read three different geometric features off the
same log-likelihood curve. The \engterm{Wald test} measures the
horizontal distance between the unrestricted MLE and the null value,
scaled by the curvature at the maximum:
\[
W = (\hat{\theta} - \theta_{0})^{2}\, I_{n}(\hat{\theta}),
\]
which is the square of the familiar estimate-over-standard-error
statistic, since $I_{n}(\hat{\theta})^{-1}$ is the squared standard
error. The \engterm{score test}, also called the Lagrange-multiplier
test, measures the slope of the log-likelihood at the null value,
scaled by the information there:
\[
S = \frac{U(\theta_{0})^{2}}{I_{n}(\theta_{0})},
\qquad U(\theta_{0}) = \pdv{\ell}{\theta}\Big|_{\theta_{0}}.
\]
The \engterm{likelihood-ratio test} measures the vertical drop in the
log-likelihood from the peak to the null,
\[
L = -2\bigl[\ell(\theta_{0}) - \ell(\hat{\theta})\bigr].
\]
All three converge in distribution to $\chi^{2}_{1}$ under $H_{0}$,
and all three agree to leading order in a neighbourhood of the
maximum, where the log-likelihood is well approximated by a
parabola. Figure \ref{fig:vol02:ch14:trinity-tests} draws the three
features on one curve. The working differences are operational. The
Wald test needs only the unrestricted fit, which is why software
reports it by default, but it is not invariant under
reparametrisation: testing $\beta = 0$ and testing $e^{\beta} = 1$
give different Wald statistics on the same data, a defect that
produces the well-known \engterm{Hauck-Donner effect} of a Wald
statistic that paradoxically shrinks as the estimate moves further
from the null in a logistic fit. The score test needs only the
restricted fit, which is its advantage when the unrestricted model is
expensive to fit or fails to converge. The likelihood-ratio test
needs both fits and is invariant under reparametrisation, which is
the reason to prefer it as the referee when the three disagree.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-trinity-tests.tex}

\paragraph{Worked example: the three statistics on a Poisson rate.}
A reliability team observes $\sum x_{i} = 18$ faults over $n = 12$
equal monitoring weeks and tests $H_{0}: \lambda = 1.0$ fault per
week against the two-sided alternative. The MLE is $\hat{\lambda} =
\bar{x} = 1.5$. The Poisson per-observation information is
$I_{1}(\lambda) = 1/\lambda$, so $I_{n}(\lambda) = n/\lambda$. The
Wald statistic uses the information at the MLE: $W = (1.5 -
1.0)^{2}\, (12/1.5) = 0.25 \cdot 8.0 = 2.00$. The score statistic
uses the information at the null: the score is $U(\lambda_{0}) =
(\sum x_{i})/\lambda_{0} - n = 18/1.0 - 12 = 6$, and $I_{n}(1.0) =
12/1.0 = 12$, so $S = 36/12 = 3.00$. The likelihood-ratio statistic
uses both fits: $\ell(\lambda) = (\sum x_{i})\ln\lambda - n\lambda$
up to a constant, so $L = -2[\ell(1.0) - \ell(1.5)] = -2[(18\ln 1.0
- 12) - (18\ln 1.5 - 18)] = -2[(0 - 12) - (7.30 - 18)] = -2[-12 +
10.70] = -2(-1.30) = 2.60$. The three statistics, $W = 2.00$,
$L = 2.60$, $S = 3.00$, bracket each other and all fall below
$\chi^{2}_{0.95,1} = 3.84$, so none rejects at $\alpha = 0.05$; the
$12$-week record does not establish a rate above one fault per week.
The spread between $2.00$ and $3.00$ is the small-sample
disagreement the asymptotic theory smooths over, and it is precisely
in this near-threshold band that the choice of statistic can flip a
decision. The engineer who reports all three, or who reports the
likelihood-ratio statistic by default, is insulated against a
reparametrisation accident the Wald statistic alone would hide.

\subsection{The delta method}

The engineer rarely wants the parameter the model is written in. A
fit returns $\hat{\lambda}$, but the design needs the mean lifetime
$1/\hat{\lambda}$; a fit returns log-odds, but the report needs a
probability. The \engterm{delta method} propagates the standard error
of an estimate through a smooth nonlinear transformation
$g(\cdot)$. If $\hat{\theta}$ is asymptotically normal with
$\sqrt{n}(\hat{\theta} - \theta) \xrightarrow{d} \mathcal{N}(0, V)$,
then a first-order Taylor expansion of $g$ about $\theta$ gives
\[
\sqrt{n}\bigl(g(\hat{\theta}) - g(\theta)\bigr)
\xrightarrow{d} \mathcal{N}\!\left(0,\, g'(\theta)^{2}\, V\right),
\]
so the standard error of $g(\hat{\theta})$ is
$\abs{g'(\hat{\theta})}$ times the standard error of $\hat{\theta}$.
The vector form replaces the scalar derivative with the gradient and
the scalar variance with the covariance matrix:
$\Var(g(\hat{\vect{\theta}})) \approx \nabla g^{\transpose}
\Sigma\, \nabla g$. The method is the chain rule applied to
uncertainty, and it is the standard tool for an error bar on any
quantity that is a function of fitted parameters.

\paragraph{Worked example: the standard error of a mean lifetime.}
The exponential fit earlier in this section returned $\hat{\lambda}
= 0.741$ per thousand hours with asymptotic standard error
$0.234$. The design needs the mean lifetime $g(\lambda) = 1/\lambda$.
The derivative is $g'(\lambda) = -1/\lambda^{2}$, so $\abs{g'
(\hat{\lambda})} = 1/0.741^{2} = 1.821$, and the delta-method
standard error of the mean lifetime is $1.821 \cdot 0.234 = 0.426$
thousand hours, on a point estimate of $1/\hat{\lambda} = 1.35$. The
symmetric delta-method interval $1.35 \pm 1.96 \cdot 0.426 = [0.52,
2.18]$ thousand hours is wider on the low side than the
endpoint-inversion interval $[0.83, 3.55]$ computed earlier, and the
two disagree because the delta method linearises a transformation
that is markedly curved at small $n$. The working rule is direct:
the delta method is the fast first answer, valid when the
transformation is gentle over the span of one standard error, and
the engineer who finds the delta-method interval disagreeing with an
endpoint-inversion or a bootstrap interval has been told that the
linearisation is straining and should report the interval that does
not assume a straight transformation.

\subsection{When the MLE fails}

The MLE fails when the regularity conditions fail. Three working
cases recur. When the support depends on the parameter (the uniform
case), the asymptotic theorems do not apply and the standard error
must be derived directly. When the parameter sits on the boundary
of $\Theta$ (a variance component pinned at zero in a hierarchical
model), the asymptotic distribution becomes a mixture and is no
longer Gaussian. When the model is misspecified (the data do not
come from the assumed family), the MLE converges to the parameter
that minimises the Kullback-Leibler divergence from the truth to
the model, not to anything the engineer would call ``the true
parameter.'' The third failure is the origin of the
\engterm{quasi-likelihood} corrections used in robust statistics
and is the practical reason an engineer should never report an MLE
without a residual diagnostic.

\begin{mastery}
When the model is misspecified the naive standard error, read off
the inverse Fisher information, is wrong, and the correction is the
\engterm{sandwich estimator}. Write the expected negative Hessian as
$A(\theta) = -\E[\partial^{2}\ell/\partial\theta^{2}]$ and the
variance of the score as $B(\theta) = \Var[\partial\ell/\partial
\theta]$. Under a correctly specified model the information identity
gives $A = B$, the two cancel, and the asymptotic variance collapses
to $A^{-1} = I(\theta)^{-1}$, the familiar Cram\'er-Rao form. Under
misspecification $A \neq B$, and the correct asymptotic variance is
the sandwich
\[
\Var(\hat{\theta}) \approx A^{-1} B\, A^{-1},
\]
estimated by plugging in the observed Hessian and the empirical
outer product of the per-observation scores. The robust or
``Huber-White'' standard errors that statistical software reports
under the heading of heteroscedasticity-consistent covariance are
exactly this sandwich applied to the regression score. The working
rule is to compare the model-based and sandwich standard errors: a
large discrepancy is itself a misspecification diagnostic, telling
the engineer that the assumed error structure does not match the
data even before a residual plot is drawn.
\end{mastery}

\begin{mastery}
The bias-variance trade-off has a counterintuitive consequence that
sharpens the case for shrinkage. Estimating three or more
independent Gaussian means simultaneously, the obvious estimator,
each sample mean estimating its own population mean, is
\engterm{inadmissible}: the \engterm{James-Stein estimator}
\[
\hat{\mu}^{\mathrm{JS}}_{i} = \left(1 -
\frac{(k-2)\sigma^{2}}{\norm{\bar{X}}^{2}}\right)\bar{X}_{i},
\qquad k \geq 3,
\]
which pulls every sample mean toward zero (or toward a common grand
mean) by a data-determined factor, has strictly smaller total mean
squared error than the vector of separate sample means, for every
true value of the means. The shrinkage factor is large when the
observed means are close together (the pooled estimate is trusted)
and small when they are spread out (the data override the pull). The
result is the frequentist face of the Bayesian hierarchical model:
shrinking many related estimates toward a common centre trades a
little bias for a large variance reduction across the ensemble, and
the trade is favourable whenever three or more quantities are
estimated together. The engineer who calibrates fifty sensors,
estimates defect rates across many production lines, or fits
per-group effects in a multilevel model is in the regime where
shrinkage beats the separate estimates, which is the practical
reason hierarchical models dominate when the groups are many and the
per-group data are thin.
\end{mastery}

\begin{mastery}
The reason the same handful of summaries, the sample mean, the sample
sum, the sum of squares, the sample maximum, recur as the estimators
for distribution after distribution is \engterm{sufficiency}. A
statistic $T(X)$ is sufficient for $\theta$ when the conditional
distribution of the data given $T$ does not depend on $\theta$: once
$T$ is known, the rest of the data carry no further information about
the parameter. The \engterm{factorisation theorem} makes this
checkable without computing a conditional distribution: $T$ is
sufficient if and only if the likelihood factors as $p(x; \theta) =
h(x)\, g(T(x); \theta)$, a part that does not involve $\theta$ times
a part that touches the data only through $T$. The \engterm{exponential
family} is the class of distributions whose density has the canonical
form
\[
p(x; \theta) = h(x)\, \exp\!\bigl(\eta(\theta)^{\transpose} T(x)
- A(\theta)\bigr),
\]
and the Gaussian, Bernoulli, Poisson, exponential, Gamma, and Beta
are all members. For an exponential family the statistic $T(x)$ in
the exponent is sufficient by inspection, which is why the sample sum
is sufficient for a Poisson rate, the count of successes is
sufficient for a Bernoulli probability, and the pair (sample sum,
sum of squares) is sufficient for a Gaussian mean and variance. The
log-partition function $A(\theta)$ carries the moments: its gradient
is the mean of $T$ and its Hessian is the covariance of $T$, which is
the Fisher information in the canonical parametrisation. The working
payoff is the \engterm{Rao-Blackwell theorem}: any estimator can be
improved, in mean squared error, by conditioning it on a sufficient
statistic, so an estimator that is not a function of the sufficient
statistic is throwing information away. The engineer who notices that
a proposed estimator ignores part of a sufficient statistic has found
a strictly better estimator without any new data.
\end{mastery}

\section{Bayesian inference}\pathtag{standard}

The Bayesian framing treats the parameter as a random variable with
its own distribution, and the data update that distribution
\cite[ch.~11]{text:wasserman2004}. Volume I Chapter 8 introduced
this on a contingency table; this section makes it formal.

\subsection{Prior, likelihood, posterior}

The engineer specifies a \engterm{prior density} $\pi(\theta)$ on
$\Theta$, expressing the state of knowledge before the data arrive.
The data, with likelihood $L(\theta) = p(x_{1}, \ldots, x_{n};
\theta)$, update the prior to the \engterm{posterior density} via
Bayes' rule:
\[
\pi(\theta \mid x) = \frac{L(\theta)\, \pi(\theta)}
                          {\int_{\Theta} L(\vartheta)\, \pi(\vartheta)\,\dd\vartheta}.
\]
The denominator, the \engterm{marginal likelihood} or
\engterm{evidence} $m(x)$, is the normalising constant that makes
$\pi(\cdot \mid x)$ integrate to one. In closed-form cases the
denominator can be ignored: the posterior is proportional to
likelihood times prior.

\subsection{Conjugate priors}

A prior is \engterm{conjugate} to a likelihood family if the
posterior belongs to the same parametric family as the prior. The
working catalogue covers the canonical models:

\textit{Beta-Bernoulli}. For $X_{i} \sim \mathrm{Bernoulli}(p)$ with
prior $p \sim \mathrm{Beta}(\alpha, \beta)$, the posterior is
$p \mid x \sim \mathrm{Beta}(\alpha + s,\, \beta + n - s)$, where
$s = \sum x_{i}$.

\textit{Gamma-Poisson}. For $X_{i} \sim \mathrm{Poisson}(\lambda)$
with prior $\lambda \sim \mathrm{Gamma}(\alpha, \beta)$, the
posterior is $\lambda \mid x \sim \mathrm{Gamma}(\alpha + s,\,
\beta + n)$.

\textit{Gaussian-Gaussian, known variance}. For $X_{i} \sim
\mathcal{N}(\mu, \sigma^{2})$ with $\sigma^{2}$ known and prior
$\mu \sim \mathcal{N}(\mu_{0}, \tau_{0}^{2})$, the posterior is
$\mu \mid x \sim \mathcal{N}(\mu_{n}, \tau_{n}^{2})$ with
\[
\tau_{n}^{-2} = \tau_{0}^{-2} + n\sigma^{-2},
\qquad
\mu_{n} = \tau_{n}^{2}\!\left(\tau_{0}^{-2}\mu_{0} + n\sigma^{-2}\bar{x}\right).
\]
The posterior precision is the sum of prior and data precisions.
The posterior mean is the precision-weighted average of the prior
mean and the sample mean. As $n$ grows, the data dominate and the
posterior collapses on the MLE.

\paragraph{Worked example: a defect rate updated batch by batch.}
A supplier's historical defect rate is about $2\,\%$, and the
engineer encodes this as a $\mathrm{Beta}(2, 98)$ prior on $p$,
with prior mean $0.02$ and effective sample size $\alpha + \beta =
100$. A first batch of $n_{1} = 50$ units has $s_{1} = 3$ defects.
The posterior after batch one is $\mathrm{Beta}(2 + 3, 98 + 47) =
\mathrm{Beta}(5, 145)$, with posterior mean $5/150 = 0.0333$. A
second batch of $n_{2} = 50$ has $s_{2} = 4$ defects; because the
Beta-Bernoulli update is sequential, the engineer can update the
posterior from batch one or, equivalently, update the original
prior with the pooled counts. Either route gives
$\mathrm{Beta}(2 + 7, 98 + 91) = \mathrm{Beta}(9, 189)$, posterior
mean $9/198 = 0.0455$. The sequential coherence of the Bayesian
update, that processing the batches one at a time yields the same
posterior as processing them together, is the property that makes
the Beta-Bernoulli model the standard engine for online
quality-control monitoring. The posterior probability that the
true defect rate exceeds a $5\,\%$ action threshold is
$\Pr[p > 0.05 \mid \text{data}]$ under $\mathrm{Beta}(9, 189)$,
which numerical integration places near $0.37$; the engineer who
has set a halt rule at posterior probability $0.5$ does not halt,
but the rising trend across batches is the signal to watch. Figure
\ref{fig:vol02:ch14:beta-posterior-update} draws the prior and the two
sequential posteriors on one axis, and the visual signal the engineer
watches is the posterior mode marching rightward toward the action
threshold as each batch's evidence sharpens and shifts the density.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-beta-posterior-update.tex}

\paragraph{Worked example: a Gaussian mean with an informative
prior.} A thermocouple's bias is believed, from the calibration
certificate, to be near zero with standard deviation
$0.5\,\degC$, encoded as $\mu \sim \mathcal{N}(0, 0.25)$. Five new
readings of a reference cell at $0\,\degC$ give a sample mean
$\bar{x} = 0.4\,\degC$ with known measurement noise $\sigma = 1.0\,
\degC$, so $\sigma^{2} = 1.0$. The prior precision is
$\tau_{0}^{-2} = 1/0.25 = 4$ and the data precision is
$n\sigma^{-2} = 5/1 = 5$. The posterior precision is
$\tau_{n}^{-2} = 4 + 5 = 9$, so $\tau_{n}^{2} = 0.111$ and the
posterior standard deviation is $0.333\,\degC$. The posterior mean
is $\tau_{n}^{2}(\tau_{0}^{-2}\mu_{0} + n\sigma^{-2}\bar{x}) =
0.111(4 \cdot 0 + 5 \cdot 0.4) = 0.111 \cdot 2.0 = 0.222\,\degC$.
The posterior pulls the data's $0.4$ toward the prior's $0$ by a
fraction $4/9 = 44\,\%$, exactly the prior's share of the total
precision. A $95\,\%$ credible interval is $0.222 \pm 1.96 \cdot
0.333 = [-0.43, 0.88]\,\degC$, which the engineer reports as a
direct probability statement about the bias, available because the
prior was a defensible certificate rather than a fabricated
convenience.

\subsection{Posterior summaries and credible intervals}

A \engterm{credible interval} of level $1-\alpha$ is an interval
$[L, U]$ such that $\Pr[\theta \in [L, U] \mid x] = 1 - \alpha$.
The probability is a posterior probability about the parameter,
given the data. The two common choices are the
\engterm{equal-tailed credible interval}, with $\alpha/2$ posterior
mass on each side, and the \engterm{highest-posterior-density
interval}, the shortest interval with total posterior mass
$1 - \alpha$. For a unimodal symmetric posterior the two coincide.

The frequentist confidence interval and the Bayesian credible
interval answer different questions. The first is a procedural
guarantee about coverage across resamples; the second is a posterior
probability statement about the parameter. With weak priors and
large samples they often coincide numerically.

\subsection{Predictive distribution}

The Bayesian's prediction for a new observation $X_{\mathrm{new}}$
integrates over the posterior:
\[
p(x_{\mathrm{new}} \mid x) = \int_{\Theta} p(x_{\mathrm{new}}; \theta)\,
                                            \pi(\theta \mid x)\,\dd\theta.
\]
The integral propagates parameter uncertainty into the prediction
in a single quantity. For the Beta-Bernoulli model, the predictive
probability of a future success is $(\alpha + s)/(\alpha + \beta +
n)$, the posterior mean of $p$. For the Gaussian-Gaussian model with
known variance, the predictive distribution is Gaussian with mean
$\mu_{n}$ and variance $\sigma^{2} + \tau_{n}^{2}$. The added
$\tau_{n}^{2}$ is the parameter uncertainty that a frequentist
prediction interval has to account for separately.

\paragraph{Worked example: a Gamma-Poisson predictive for counts.}
A maintenance team logs equipment faults per week and models the
weekly count as $X_{i} \sim \mathrm{Poisson}(\lambda)$ with a
conjugate $\mathrm{Gamma}(\alpha, \beta)$ prior on the rate, where
$\beta$ is the rate parameter of the Gamma so the prior mean is
$\alpha/\beta$. From two years of comparable equipment the team sets
$\alpha = 6$, $\beta = 3$, a prior mean of $2$ faults per week with
the inferential weight of $\beta = 3$ prior weeks. Observing eight
weeks with counts summing to $s = 22$ gives the posterior
$\mathrm{Gamma}(\alpha + s, \beta + n) = \mathrm{Gamma}(28, 11)$,
posterior mean $28/11 = 2.55$ faults per week. The predictive
distribution for next week's count integrates the Poisson over the
Gamma posterior, and the integral is a \engterm{negative binomial}:
\[
p(x_{\mathrm{new}} \mid x)
= \int_{0}^{\infty} \frac{\lambda^{x_{\mathrm{new}}}e^{-\lambda}}
{x_{\mathrm{new}}!}\, \frac{\beta_{n}^{\alpha_{n}}}{\Gamma(\alpha_{n})}
\lambda^{\alpha_{n}-1}e^{-\beta_{n}\lambda}\,\dd\lambda
= \binom{x_{\mathrm{new}} + \alpha_{n} - 1}{x_{\mathrm{new}}}
\left(\frac{\beta_{n}}{\beta_{n}+1}\right)^{\alpha_{n}}
\left(\frac{1}{\beta_{n}+1}\right)^{x_{\mathrm{new}}},
\]
with $\alpha_{n} = 28$ and $\beta_{n} = 11$. Its mean is
$\alpha_{n}/\beta_{n} = 2.55$, matching the posterior mean of
$\lambda$, but its variance is $\alpha_{n}(\beta_{n}+1)/\beta_{n}^{2}
= 28 \cdot 12/121 = 2.78$, larger than the bare Poisson variance of
$2.55$. The excess, the \engterm{overdispersion}, is the parameter
uncertainty propagated into the prediction: a maintenance planner
who sizes spare-part stock against the negative-binomial predictive
holds slightly more than a planner who fixes $\lambda$ at its
posterior mean and uses a bare Poisson, and the difference is the
honest cost of not knowing $\lambda$ exactly. The predictive
probability of a quiet week with zero faults is
$(\beta_{n}/(\beta_{n}+1))^{\alpha_{n}} = (11/12)^{28} = 0.087$,
and the probability of five or more faults, the tail a planner
worries about, is about $0.10$ under the predictive against $0.07$
under the plug-in Poisson, a $40\,\%$ relative underestimate of the
busy-week risk that the plug-in commits.

\subsection{A Bayesian decision under a loss function}

A posterior is not yet a decision. The Bayesian framing closes only
when the engineer attaches a \engterm{loss function} and chooses the
action that minimises expected posterior loss. The pattern is the
decision-theoretic framing of section 14.2 carried through to a
posterior rather than a sampling distribution, and it is the right
machine for the A/B choice that recurs whenever two process settings,
two suppliers, or two designs are compared on a binary outcome.

\paragraph{Worked example: an A/B decision with an asymmetric loss.}
Two coating processes are compared on the probability that a part
passes a corrosion test. Process A, the incumbent, has a long record
encoded as a $\mathrm{Beta}(40, 10)$ posterior, mean pass rate
$0.80$. Process B, a candidate, has been run on $30$ parts with $27$
passes; under a flat $\mathrm{Beta}(1, 1)$ prior its posterior is
$\mathrm{Beta}(28, 4)$, mean $0.875$. The engineer wants the
probability that B is genuinely better, $\Pr[p_{B} > p_{A} \mid
\text{data}]$, which is a two-dimensional posterior integral over the
product of the two independent Beta posteriors. Monte Carlo settles
it directly: draw $p_{A}$ from $\mathrm{Beta}(40, 10)$ and $p_{B}$
from $\mathrm{Beta}(28, 4)$ ten thousand times and count the fraction
with $p_{B} > p_{A}$, which lands near $0.88$. The point estimate
favours B, and the posterior probability that B is better is high but
not decisive.

The decision needs the cost structure, not the probability alone.
Switching to B carries a one-time changeover cost equivalent to a
$0.02$ drop in pass rate amortised over the production run, and
staying with A forgoes whatever real improvement B offers. Encode the
loss of choosing B as $\ell_{B} = \max(0,\, p_{A} - p_{B}) + 0.02$,
the regret if A was actually better plus the changeover penalty, and
the loss of choosing A as $\ell_{A} = \max(0,\, p_{B} - p_{A})$, the
regret of forgoing B's improvement. Averaging each loss over the same
Monte Carlo draws gives an expected loss of staying with A of about
$0.078$ (dominated by the $0.075$ mean improvement B offers when B is
better) and an expected loss of switching to B of about $0.025$ (the
$0.02$ changeover plus the small probability-weighted regret that A
was better). The expected-loss-minimising action is to switch, and
the margin, $0.078$ against $0.025$, is the quantity the engineer
reports rather than the bare posterior probability. Had the
changeover cost been equivalent to a $0.08$ pass-rate drop instead of
$0.02$, the expected loss of switching would have risen to about
$0.085$, exceeding the cost of staying, and the same posterior would
have selected the opposite action. The lesson is that the posterior
is an input to the decision, and the loss function is the other
input; an A/B test reported as ``B wins, posterior probability
$0.88$'' has reported half the analysis.

\subsection{When to reach for Bayesian inference}

The engineer reaches for Bayesian inference when prior information
is genuine and quantified, when the parameter has a structured
hierarchical interpretation, when the goal is a posterior over many
parameters with explicit uncertainty propagation, or when the data
are sparse enough that classical asymptotics fail. The engineer
declines Bayesian inference when the prior is fabricated,
high-dimensional integration is computationally infeasible without
machinery the project does not justify, or the audience for the
result requires a frequentist procedural guarantee that the
posterior credible interval does not provide.

\begin{mastery}
The choice of prior is the load-bearing modelling step. A
\engterm{non-informative} or \engterm{weakly informative} prior
attempts to let the data speak. Jeffreys' prior, $\pi_{J}(\theta)
\propto \sqrt{\det I(\theta)}$, is invariant under reparametrisation
and is a defensible default for many one-parameter models. A
\engterm{conjugate prior with small effective sample size} (small
$\alpha + \beta$ in a Beta prior, small $\tau_{0}^{-2}$ in a
Gaussian prior) is the engineer's working compromise: it stabilises
the posterior in small-sample regimes while letting the data
dominate as $n$ grows. A \engterm{strong prior} reflects genuine
knowledge from earlier studies or physical constraints, and must be
defensible as such; otherwise it is wishful thinking.
\end{mastery}

\begin{mastery}
The Bayesian comparison of two models is the ratio of their
evidences, the \engterm{Bayes factor}. For models $M_{1}$ and
$M_{2}$ with marginal likelihoods $m_{k}(x) = \int p(x \mid \theta,
M_{k})\, \pi(\theta \mid M_{k})\, \dd\theta$, the Bayes factor is
\[
B_{12} = \frac{m_{1}(x)}{m_{2}(x)}
= \frac{\Pr[M_{1} \mid x]/\Pr[M_{2} \mid x]}
       {\Pr[M_{1}]/\Pr[M_{2}]},
\]
the factor by which the data move the prior odds on the two models to
the posterior odds. A Bayes factor of $10$ means the data are ten
times as probable under $M_{1}$ as under $M_{2}$; the conventional
reading places $B_{12}$ above $3$ as positive evidence, above $20$ as
strong, and above $150$ as very strong, with the symmetric brackets
below one favouring $M_{2}$ \cite{text:wasserman2004}. The Bayes
factor builds in an automatic penalty for complexity, the
\engterm{Bayesian Occam factor}: a model with more free parameters
spreads its prior predictive mass over a larger data space, so it
assigns less probability to any particular observed dataset, and the
marginal likelihood is lower unless the extra parameters earn their
keep. This is the Bayesian counterpart to cross-validation's penalty
on flexibility, reached by integration rather than by holding out
data. The working caution is that the marginal likelihood depends on
the prior in a way the posterior does not: a diffuse prior on a
parameter that the data constrain well still suppresses the evidence,
which is the \engterm{Lindley-Jeffreys paradox}, where a frequentist
test rejects the null while a Bayes factor under a vague prior favours
it. The engineer who reports a Bayes factor reports the prior it was
computed under, because a Bayes factor without its prior is a number
without a referent, and the approximation by the Bayesian information
criterion, $-2\ln B_{12} \approx \mathrm{BIC}_{1} - \mathrm{BIC}_{2}$,
is the cheap large-sample stand-in when the integral is intractable.
\end{mastery}

\section{Regression: linear, logistic, regularised}\pathtag{core}

Volume I Chapter 8 fitted a straight line to data on trust. Volume
II Chapter 9 supplied the linear-algebraic apparatus, including the
projection geometry and the conditioning analysis. This section
combines them into a statistical model and extends the framework to
binary outcomes and to penalised estimation.

\subsection{The linear regression model}

The data are $(x_{i}, y_{i})$ for $i = 1, \ldots, n$, with $x_{i}
\in \R^{p}$ a vector of predictors and $y_{i} \in \R$ a response.
The \engterm{linear regression model} is
\[
Y_{i} = x_{i}^{\transpose}\beta + \varepsilon_{i},
\qquad \varepsilon_{i} \sim \mathcal{N}(0, \sigma^{2}),
\quad \text{independent across } i.
\]
Stacking gives $\vect{Y} = \mat{X}\vect{\beta} + \vect{\varepsilon}$,
where $\mat{X}$ is the \engterm{design matrix} of size $n \times p$
with rows $x_{i}^{\transpose}$. The Gaussian noise assumption
identifies the model: the likelihood is
\[
L(\vect{\beta}, \sigma^{2})
= \prod_{i=1}^{n}\frac{1}{\sqrt{2\pi\sigma^{2}}}
  \exp\!\left(-\frac{(y_{i} - x_{i}^{\transpose}\vect{\beta})^{2}}{2\sigma^{2}}\right),
\]
and maximising the likelihood in $\vect{\beta}$ is equivalent to
minimising the residual sum of squares.

\subsection{Least squares as MLE under Gaussian noise}

The MLE for $\vect{\beta}$ is the \engterm{ordinary least squares
estimator},
\[
\hat{\vect{\beta}}_{\mathrm{OLS}}
  = \argmin_{\vect{\beta} \in \R^{p}}
    \norm*{\vect{Y} - \mat{X}\vect{\beta}}_{2}^{2}.
\]
Volume II Chapter 9 derived the closed form by setting the
gradient with respect to $\vect{\beta}$ to zero, producing the
\engterm{normal equations} $\mat{X}^{\transpose}\mat{X}
\hat{\vect{\beta}} = \mat{X}^{\transpose}\vect{Y}$. When $\mat{X}$
has full column rank, the solution is
\[
\hat{\vect{\beta}}_{\mathrm{OLS}}
  = (\mat{X}^{\transpose}\mat{X})^{-1}\mat{X}^{\transpose}\vect{Y}.
\]
The estimator is unbiased: $\E[\hat{\vect{\beta}}] = \vect{\beta}$.
Its covariance is $\Var(\hat{\vect{\beta}}) = \sigma^{2}
(\mat{X}^{\transpose}\mat{X})^{-1}$. The Gauss-Markov theorem
states that, under the Gaussian-noise model, the OLS estimator has
the smallest variance among all linear unbiased estimators.

\subsection{Diagnostic plots and residual analysis}

The MLE is unreliable without residual diagnostics. The
\engterm{residual} for observation $i$ is $\hat{\varepsilon}_{i} =
y_{i} - x_{i}^{\transpose}\hat{\vect{\beta}}$. Four plots carry
most of the diagnostic load:

\begin{itemize}[leftmargin=*]
  \item Residuals against fitted values $\hat{y}_{i}$. A horizontal
    band of constant width is consistent with the homoscedastic
    Gaussian model. A funnel shape signals heteroscedasticity. A
    curve signals a missing nonlinearity.
  \item Residuals against each predictor. A pattern signals the
    same missing nonlinearity, localised to one variable.
  \item A normal quantile-quantile plot of the residuals. Severe
    departures in the tails signal that the Gaussian-noise
    assumption is wrong.
  \item Leverage and Cook's distance, to flag observations whose
    removal would change the fit substantially.
\end{itemize}

A regression report without diagnostic plots is a regression report
the engineer should not believe.

\paragraph{A fully worked least-squares fit.} Eight tensile
specimens are tested; the predictor $x$ is the carbon content in
hundredths of a percent and the response $y$ is the yield strength
in arbitrary working units. The data are
\[
\begin{array}{c|cccccccc}
x & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\ \hline
y & 3.1 & 4.0 & 5.2 & 5.8 & 7.1 & 7.6 & 9.0 & 9.2
\end{array}
\]
The sample means are $\bar{x} = 4.5$ and $\bar{y} = 6.375$. The
centred sums are $S_{xx} = \sum(x_{i} - \bar{x})^{2} = 42$ and
$S_{xy} = \sum(x_{i} - \bar{x})(y_{i} - \bar{y}) = 38.10$. The slope
is $\hat{\beta}_{1} = S_{xy}/S_{xx} = 38.10/42 = 0.907$ and the
intercept is $\hat{\beta}_{0} = \bar{y} - \hat{\beta}_{1}\bar{x} =
6.375 - 0.907 \cdot 4.5 = 2.293$. The fitted line is
$\hat{y} = 2.293 + 0.907 x$.

The residuals $\hat{\varepsilon}_{i} = y_{i} - \hat{y}_{i}$ are
\[
\begin{array}{c|cccccccc}
x & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\ \hline
\hat{y} & 3.20 & 4.11 & 5.01 & 5.92 & 6.83 & 7.74 & 8.64 & 9.55 \\
\hat{\varepsilon} & -0.10 & -0.11 & 0.19 & -0.12 & 0.27 & -0.14 & 0.36 & -0.35
\end{array}
\]
The residual sum of squares is $\mathrm{RSS} = \sum
\hat{\varepsilon}_{i}^{2} = 0.413$, so the residual variance
estimate is $\hat{\sigma}^{2} = \mathrm{RSS}/(n - 2) = 0.413/6 =
0.0688$ and $\hat{\sigma} = 0.262$. The standard error of the slope
is $\hat{\sigma}/\sqrt{S_{xx}} = 0.262/\sqrt{42} = 0.0405$, giving
a $95\,\%$ confidence interval for the slope of $0.907 \pm
t_{0.975, 6} \cdot 0.0405 = 0.907 \pm 2.447 \cdot 0.0405 =
[0.808, 1.006]$. The coefficient of determination is $R^{2} = 1 -
\mathrm{RSS}/S_{yy}$; with $S_{yy} = \sum(y_{i} - \bar{y})^{2} =
35.06$, $R^{2} = 1 - 0.413/35.06 = 0.988$. The fit is tight, the
slope interval excludes zero by a wide margin, and the residuals
show no obvious pattern against $x$, so the linear model is
defensible on these data. Figure
\ref{fig:vol02:ch14:regression-residuals} plots the fit and the
residuals together; the residual panel is where an engineer looks
first, because a high $R^{2}$ with a curved residual band is a
fitted line through a curved truth.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-regression-residuals.tex}

\begin{estimation}
\textbf{Question.} An engineer plans a regression to estimate a
slope $\beta_{1}$ to a precision of $\pm 0.05$ (the half-width of
the $95\,\%$ confidence interval), with predictors spread over a
range of about $10$ units and a working residual standard deviation
of $\hat{\sigma} \approx 0.3$. About how many evenly spaced
observations are needed?

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The slope standard error is $\hat{\sigma}/\sqrt{S_{xx}}$, and the
half-width is roughly $2\,\hat{\sigma}/\sqrt{S_{xx}}$ at large $n$.
For $n$ points evenly spread over a range $R$, the centred sum of
squares is approximately $S_{xx} \approx n R^{2}/12$. Setting
$2\,\hat{\sigma}/\sqrt{nR^{2}/12} = 0.05$ and solving gives
$\sqrt{n} = 2\,\hat{\sigma}\sqrt{12}/(0.05\,R) = 2 \cdot 0.3 \cdot
3.46/(0.05 \cdot 10) = 2.08/0.5 = 4.16$, so $n \approx 17$. The
working lesson is that slope precision improves with both the
sample size and the spread of the predictor: doubling the predictor
range $R$ buys the same precision as quadrupling $n$, because
$S_{xx}$ scales with $R^{2}$. An engineer who can choose the design
points should spread them to the edges of the defensible domain,
not cluster them at the centre.
\end{estimation}

\subsection{Logistic regression}

When the response is binary, $y_{i} \in \set*{0, 1}$, the linear
model is replaced by the \engterm{logistic regression model}:
\[
\Pr[Y_{i} = 1 \mid x_{i}] = \sigma(x_{i}^{\transpose}\vect{\beta}),
\qquad \sigma(z) = \frac{1}{1 + e^{-z}}.
\]
The function $\sigma$ is the logistic sigmoid; its inverse $\log
(p/(1-p))$ is the \engterm{logit} or log-odds. The likelihood is
\[
L(\vect{\beta}) = \prod_{i=1}^{n}
   \sigma(x_{i}^{\transpose}\vect{\beta})^{y_{i}}\,
   (1 - \sigma(x_{i}^{\transpose}\vect{\beta}))^{1 - y_{i}},
\]
and the MLE has no closed form. The standard fitting routine is
\engterm{iteratively reweighted least squares}, equivalent to
Newton's method on the log-likelihood. Volume II Chapter 15 gives
the optimisation apparatus; this chapter takes it for granted.

A logistic regression coefficient $\beta_{j}$ is interpreted on the
log-odds scale: a unit change in $x_{j}$ multiplies the odds
$p/(1-p)$ by $e^{\beta_{j}}$. The interpretation does not match the
linear-regression interpretation, and an engineer who reads
$\beta_{j}$ as a probability change has misread the model.

\paragraph{Worked example: reading a fitted logistic model.} A
fitted single-predictor model for the probability that a weld
passes inspection, as a function of travel speed $x$ in mm/s, is
$\hat{\beta}_{0} = 4.0$ and $\hat{\beta}_{1} = -0.5$. The fitted
log-odds at $x = 6$ is $4.0 - 0.5 \cdot 6 = 1.0$, so the fitted
pass probability is $\sigma(1.0) = 1/(1 + e^{-1.0}) = 0.731$. At
$x = 8$ the log-odds is $0$ and the pass probability is exactly
$0.5$; the predictor value at which $\hat{p} = 0.5$ is the
\engterm{median-effective level}, here $x = -\hat{\beta}_{0}/
\hat{\beta}_{1} = 8\,$mm/s. The coefficient $\hat{\beta}_{1} =
-0.5$ means a one-unit increase in travel speed multiplies the
pass odds by $e^{-0.5} = 0.607$, a $39\,\%$ reduction in odds per
mm/s. The probability change is not constant: moving from
$x = 6$ to $x = 7$ drops the pass probability from $0.731$ to
$0.622$ (a drop of $0.109$), while moving from $x = 8$ to $x = 9$
drops it from $0.500$ to $0.378$ (a drop of $0.122$). The
steepest probability change is at the median-effective level, which
is the geometric reason the odds-ratio interpretation, constant
across the range, is the one the coefficient supports and the
probability-change interpretation, varying across the range, is the
one it does not. Figure \ref{fig:vol02:ch14:logistic-fit} draws the
fitted curve through the raw pass and fail observations and marks the
median-effective level where the curve is steepest.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-logistic-fit.tex}

\paragraph{Worked example: one IRLS step by hand.} The logistic MLE
has no closed form, and the standard solver is Newton's method on
the log-likelihood, which rearranges into a weighted least-squares
update, hence \engterm{iteratively reweighted least squares}. The
update is
\[
\vect{\beta}^{(t+1)} = \left(\mat{X}^{\transpose}\mat{W}\mat{X}
\right)^{-1}\mat{X}^{\transpose}\mat{W}\vect{z},
\qquad
\vect{z} = \mat{X}\vect{\beta}^{(t)} + \mat{W}^{-1}(\vect{y} -
\vect{p}),
\]
where $\vect{p}$ holds the current fitted probabilities $p_{i} =
\sigma(x_{i}^{\transpose}\vect{\beta}^{(t)})$, $\mat{W} =
\operatorname{diag}(p_{i}(1-p_{i}))$ is the diagonal weight matrix,
and $\vect{z}$ is the \engterm{working response}. We take one step on
a four-point intercept-plus-slope model with predictor values
$x = (2, 4, 6, 8)$ and outcomes $y = (1, 1, 0, 0)$, starting from
$\vect{\beta}^{(0)} = (0, 0)^{\transpose}$. At the start every
fitted probability is $p_{i} = \sigma(0) = 0.5$, so each weight is
$p_{i}(1-p_{i}) = 0.25$ and $\mat{W} = 0.25\,\mat{I}_{4}$. The
working response is $z_{i} = 0 + (y_{i} - 0.5)/0.25 = 4(y_{i} -
0.5)$, giving $\vect{z} = (2, 2, -2, -2)^{\transpose}$. The
weighted normal equations reduce, because $\mat{W}$ is a scalar
multiple of the identity at this first step, to an ordinary
least-squares fit of $\vect{z}$ on $(1, x)$: with $\bar{x} = 5$,
$\bar{z} = 0$, $S_{xx} = (2-5)^{2} + (4-5)^{2} + (6-5)^{2} + (8-5)^{2}
= 20$, and $S_{xz} = (2-5)(2) + (4-5)(2) + (6-5)(-2) + (8-5)(-2) =
-6 - 2 - 2 - 6 = -16$, the slope is $S_{xz}/S_{xx} = -0.80$ and the
intercept is $\bar{z} - (-0.80)\bar{x} = 0 + 4.0 = 4.0$. After one
step $\vect{\beta}^{(1)} = (4.0, -0.80)^{\transpose}$, already close
to the fitted weld model's shape and headed toward the MLE that the
solver reaches in three or four further iterations. The example
exposes the mechanism behind the closed-box logistic fit: each
Newton step is a weighted regression of a working response, the
weights are the variances of the Bernoulli outcomes at the current
fit, and the iteration converges because the log-likelihood is
concave in $\vect{\beta}$.

\subsection{Applied vignette: a defect classifier and the threshold decision}

A fitted logistic model returns a probability, and a probability is not
yet a decision. This vignette carries a weld-inspection classifier from
a fitted model to a deployed decision rule, and it shows that the model
and the operating threshold are separate choices governed by separate
evidence, the point the receiver operating characteristic exists to make.

An automated inspection station photographs each weld and a logistic
model returns $\hat{p}_{i}$, the fitted probability that weld $i$ is
defective, from a handful of image features. The plant has $n = 400$
welds with known ground-truth labels, of which $60$ are genuinely
defective, a defect prevalence of $15\,\%$. The engineer must convert
each $\hat{p}_{i}$ into a keep-or-reject action by fixing a threshold
$t$: reject the weld when $\hat{p}_{i} > t$. Every threshold produces a
$2\times2$ table of outcomes. At the textbook default $t = 0.5$ the
model flags $48$ welds, of which $42$ are truly defective (true
positives) and $6$ are sound welds wrongly rejected (false positives),
while $18$ defective welds slip through (false negatives). The
\engterm{sensitivity}, the fraction of defects caught, is $42/60 =
0.70$; the \engterm{specificity}, the fraction of sound welds correctly
passed, is $334/340 = 0.982$; and the overall accuracy is $(42 +
334)/400 = 0.94$. The $94\,\%$ accuracy looks strong until the
prevalence is recalled: a trivial classifier that passes every weld
scores $340/400 = 0.85$ accuracy by doing nothing, so accuracy on an
imbalanced problem flatters a weak model. Accuracy is the wrong
headline figure for a rare-event classifier, which is why the vignette
does not report it alone.

Sweeping the threshold from strict to permissive traces the
\engterm{receiver operating characteristic}, the curve of sensitivity
against the false-positive rate $1 - \text{specificity}$ that figure
\ref{fig:vol02:ch14:roc-curve} draws. At a strict threshold near $t =
0.8$ the station rejects only the welds it is most sure of, catching few
defects but almost never rejecting a sound weld (high specificity, low
sensitivity, the lower-left of the curve). At a permissive threshold
near $t = 0.2$ it rejects liberally, catching almost every defect at the
cost of many false rejections (high sensitivity, low specificity, the
upper-right). The model fixes the shape of the curve; the engineer
chooses the point on it. The \engterm{area under the curve}, near
$0.88$ for this model, summarises the ranking quality across all
thresholds and is prevalence-free, which is why it, not accuracy,
compares two candidate models.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-roc-curve.tex}

The threshold itself is a decision-theoretic quantity, and the cost
structure sets it. A defective weld that escapes to the field carries a
warranty and safety cost the plant estimates at $C_{\mathrm{FN}} =
2000$ working units; a sound weld wrongly rejected costs a rework
$C_{\mathrm{FP}} = 50$ units. The Bayes-optimal threshold on the fitted
probability minimises expected cost and works out to $t^{\star} =
C_{\mathrm{FP}}/(C_{\mathrm{FP}} + C_{\mathrm{FN}}) = 50/2050 =
0.024$, far below the $0.5$ default: because a missed defect costs
forty times a false rejection, the station should reject on even a
small fitted probability of a defect. At that threshold the sensitivity
rises toward $0.97$ and the false-positive rate rises with it, and the
engineer accepts the extra rework as the cheaper error. The lesson is
the vignette's point: the model produces a probability, the ROC curve
shows every threshold the probability could take, and the cost ratio
selects the one threshold the plant should deploy. A vendor who reports
``$94\,\%$ accurate'' has reported neither the curve nor the threshold
nor the cost ratio, and the number is silent on the decision the plant
actually has to make.

\subsection{Applied vignette: is the probability a probability? calibration}

The ROC vignette read the classifier's output as a \emph{ranking} and
asked which threshold to cut it at. A separate question the ROC curve
cannot answer is whether the number the model reports as a probability
behaves like one. A model that assigns $0.7$ to a batch of welds should
see about seventy percent of that batch turn out defective; if instead
only forty percent do, the model is \engterm{miscalibrated}, and every
downstream decision that reads the probability literally, expected-cost
thresholds, spare-part sizing, risk pricing, inherits the error. The
distinction is sharp: ranking and calibration are independent
properties, and a model can rank cases perfectly while reporting
probabilities that are systematically too confident.

Calibration is measured directly. Bin the predicted probabilities into
intervals, and in each bin compare the mean predicted probability to
the observed frequency of the positive class. A perfectly calibrated
model places every bin on the $45$-degree line of the
\engterm{reliability diagram}. Return to the weld classifier of the ROC
vignette and read its ten deciles. The model's high-confidence bin, the
welds it calls $95\,\%$ likely defective, contains parts that are
defective only $82\,\%$ of the time; its low-confidence bin, the welds
it calls $5\,\%$ likely, contains parts defective $14\,\%$ of the time.
The reliability diagram in figure \ref{fig:vol02ch14:calibration} shows
the curve crossing the diagonal with a shallower slope than one, the
signature of an \engterm{overconfident} model that pushes its scores
toward the extremes. This defect is invisible to the ROC curve, whose
area was a healthy $0.88$: the model orders welds correctly, so every
threshold sweep looks fine, while the probabilities it hands to a
cost calculation are wrong.

The single-number summary is the \engterm{Brier score}, the mean
squared difference between the predicted probability and the binary
outcome, $\mathrm{BS} = \frac{1}{n}\sum_{i}(\hat{p}_{i} - y_{i})^{2}$.
It is a \engterm{proper scoring rule}: it is minimised in expectation
only by reporting the true probability, so a model cannot game it by
shading its forecasts toward zero or one. The Brier score decomposes,
by the Murphy decomposition, into a calibration term (how far the
reliability curve sits from the diagonal), a refinement term (how
sharply the model separates the classes), and an irreducible term set
by the base rate. A model can lower its Brier score either by
calibrating better or by sharpening, and reading the decomposition
tells the engineer which lever is available. For the weld classifier
the calibration term is the large one, which says the fix is not a
better feature set but a \engterm{recalibration} of the existing
scores: fit a one-parameter logistic map from the raw score to a
calibrated probability (\engterm{Platt scaling}) or a monotone
step function (\engterm{isotonic regression}) on a held-out
calibration set, and the reliability curve is pulled onto the diagonal
without touching the ranking the ROC curve rewarded.

The vignette's working point closes the pair. The ROC curve certifies
that the model ranks; the reliability diagram certifies that its
numbers can be read as probabilities; the two are separate audits and a
deployment that feeds probabilities into a cost model needs both. A
vendor who reports an area under the ROC curve of $0.88$ has reported
the ranking audit and stayed silent on the calibration one, and an
expected-cost threshold computed from uncalibrated scores places the
cut in the wrong place by exactly the amount the reliability diagram
would have exposed. The engineer asks for the reliability diagram and
the Brier score, not the ROC area alone, whenever the probability
itself, and not merely the ordering, drives the decision.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-calibration.tex}

\subsection{Generalised linear models and Poisson regression}

Linear regression and logistic regression are two instances of one
structure, the \engterm{generalised linear model}, which the engineer
meets again whenever the response is not a continuous Gaussian
quantity \cite[ch.~4]{text:hastie-tibshirani-friedman}. The structure
has three parts: a response distribution from the exponential family,
a \engterm{linear predictor} $\eta_{i} = x_{i}^{\transpose}\vect{\beta}$,
and a \engterm{link function} $g$ that connects the mean of the
response to the linear predictor, $g(\E[Y_{i}]) = \eta_{i}$. Ordinary
linear regression is the Gaussian response with the identity link;
logistic regression is the Bernoulli response with the logit link.
The count case is the third member the engineer needs by name.

\engterm{Poisson regression} models a count response $Y_{i} \in
\set{0, 1, 2, \ldots}$ with the log link:
\[
Y_{i} \sim \mathrm{Poisson}(\mu_{i}),
\qquad \ln\mu_{i} = x_{i}^{\transpose}\vect{\beta}.
\]
The log link guarantees the fitted mean stays positive, and it makes
the coefficients multiplicative: a unit change in $x_{j}$ multiplies
the expected count by $e^{\beta_{j}}$, the count analogue of the
logistic odds ratio. The model is the working tool for defect counts,
fault arrivals, accident rates, and any response that is a number of
events over an exposure. When the exposure varies across observations,
say a defect count over a panel of varying area $A_{i}$, the model
carries an \engterm{offset}, $\ln\mu_{i} = \ln A_{i} +
x_{i}^{\transpose}\vect{\beta}$, the term with a fixed coefficient of
one that turns the model into a rate model for events per unit
exposure.

\paragraph{Worked example: a defect-rate model with an offset.}
Painted panels are inspected for surface defects; the count $y_{i}$ of
defects is recorded along with the panel area $A_{i}$ in square metres
and a single binary predictor $x_{i}$, whether the panel was cured in
the new oven ($x = 1$) or the old one ($x = 0$). A Poisson regression
with a log-area offset returns $\hat{\beta}_{0} = -1.20$ and
$\hat{\beta}_{1} = -0.45$. The intercept sets the baseline defect rate
in the old oven: $e^{-1.20} = 0.301$ defects per square metre. The
slope multiplies that rate for the new oven by $e^{-0.45} = 0.638$, a
$36\,\%$ reduction in defect rate per unit area, so the new oven's
fitted rate is $0.301 \cdot 0.638 = 0.192$ defects per square metre.
The interpretation is a rate ratio, and the engineer who needs an
absolute defect count for a $2.5\,\si{\meter\squared}$ panel from the
new oven computes $\mu = A\, e^{\beta_{0} + \beta_{1}} = 2.5 \cdot
0.192 = 0.48$ expected defects. The standard error of the rate ratio
follows from the delta method or, more directly, from exponentiating
the endpoints of the Wald interval on $\hat{\beta}_{1}$, never by
exponentiating a symmetric interval's centre and adding the
exponentiated half-width, which would misplace the asymmetry the log
link creates.

\paragraph{Overdispersion, the Poisson model's standard failure.}
The Poisson model carries one rigid assumption, that the variance
equals the mean. Real count data routinely show variance larger than
the mean, \engterm{overdispersion}, when unmodelled heterogeneity or
clustering inflates the spread. The diagnostic is the ratio of the
residual deviance to its degrees of freedom: a ratio near one is
consistent with the Poisson model, and a ratio well above one signals
overdispersion. The fix is either a \engterm{quasi-Poisson} model,
which inflates every standard error by the square root of the
dispersion ratio while leaving the coefficients unchanged, or a
\engterm{negative-binomial} model, which adds a second parameter for
the extra variance and is the same negative binomial that arose as the
Gamma-Poisson predictive in the Bayesian section. An engineer who fits
a Poisson model and reports its standard errors without checking the
dispersion ratio has, on overdispersed data, reported intervals that
are too narrow by exactly the factor the dispersion check would have
caught.

\begin{mastery}
The Gaussian mixture and the missing-data problem behind it are
solved by the \engterm{expectation-maximisation algorithm}, the
workhorse for likelihoods that are intractable directly but easy once
a hidden label is filled in
\cite[ch.~8]{text:hastie-tibshirani-friedman}. Suppose a sample of
vibration amplitudes is a mixture of two Gaussian populations,
healthy and faulty bearings, with unknown mixing weight $\pi$, means
$\mu_{1}, \mu_{2}$, and variances $\sigma_{1}^{2}, \sigma_{2}^{2}$,
and the label of each observation is unobserved. The likelihood is a
sum of two Gaussians per observation, and its logarithm does not
separate, so direct maximisation is awkward. The algorithm alternates
two steps. The \engterm{E-step} computes, for each observation, the
posterior \engterm{responsibility} that component $k$ generated it,
\[
\gamma_{ik} = \frac{\pi_{k}\, \mathcal{N}(x_{i}; \mu_{k},
\sigma_{k}^{2})}{\sum_{j}\pi_{j}\,\mathcal{N}(x_{i}; \mu_{j},
\sigma_{j}^{2})},
\]
a soft assignment between zero and one. The \engterm{M-step} then
refits each component's parameters as responsibility-weighted sample
statistics, $\hat{\mu}_{k} = \sum_{i}\gamma_{ik}x_{i}/\sum_{i}
\gamma_{ik}$ and likewise for the variances and the weights. The two
steps repeat; each iteration is guaranteed not to decrease the
observed-data log-likelihood, which is the algorithm's defining
property and the reason it converges, though only to a local maximum,
so several random starts are the working safeguard. The pattern, fill
in the missing label by its expectation, then maximise as if it were
observed, generalises far beyond mixtures: it is the engine behind
hidden-Markov-model fitting, factor analysis with missing entries, and
censored-data likelihoods. The failure mode is its own: a component
variance can collapse toward zero around a single point, sending the
likelihood to infinity in a degenerate spike, which is why working
implementations floor the variances or place a weak prior on them.
\end{mastery}

\subsection{Regularised regression: ridge and lasso}

When the design matrix is ill-conditioned, the OLS estimator has
high variance even though it is unbiased. The bias-variance
decomposition of section 14.1 licenses a trade: introduce a small
bias to gain a large variance reduction. The two canonical
penalised estimators are ridge and lasso.

The \engterm{ridge regression} estimator is
\[
\hat{\vect{\beta}}_{\mathrm{ridge}}(\lambda)
 = \argmin_{\vect{\beta}}\!\left\{
   \norm*{\vect{Y} - \mat{X}\vect{\beta}}_{2}^{2}
 + \lambda\, \norm*{\vect{\beta}}_{2}^{2}\right\},
\]
with closed form $\hat{\vect{\beta}}_{\mathrm{ridge}}(\lambda) =
(\mat{X}^{\transpose}\mat{X} + \lambda \mat{I})^{-1}
\mat{X}^{\transpose}\vect{Y}$. The penalty $\lambda \geq 0$ is the
regularisation parameter; $\lambda = 0$ recovers OLS, and $\lambda
\to \infty$ shrinks the estimator to zero. The matrix
$\mat{X}^{\transpose}\mat{X} + \lambda\mat{I}$ has condition number
no worse than $(\sigma_{1}^{2} + \lambda)/(\sigma_{p}^{2} +
\lambda)$, which is the operational reason ridge stabilises the
solution when the design matrix is ill-conditioned.

The \engterm{lasso} replaces the squared $\ell_{2}$ penalty with an
$\ell_{1}$ penalty:
\[
\hat{\vect{\beta}}_{\mathrm{lasso}}(\lambda)
 = \argmin_{\vect{\beta}}\!\left\{
   \norm*{\vect{Y} - \mat{X}\vect{\beta}}_{2}^{2}
 + \lambda\, \norm*{\vect{\beta}}_{1}\right\},
\]
where $\norm*{\vect{\beta}}_{1} = \sum_{j} \abs*{\beta_{j}}$. The
$\ell_{1}$ geometry has corners on the coordinate axes, and the
optimum often sits at a corner: many coefficients are exactly zero
at the lasso solution. Lasso therefore performs simultaneous
shrinkage and variable selection. The closed form does not exist;
coordinate descent and proximal-gradient methods are the standard
solvers. Volume II Chapter 15 carries the optimisation; this chapter
recognises the result.

Figure \ref{fig:vol02:ch14:shrinkage-paths} draws the two penalties'
coefficient paths side by side as the penalty strength grows. Ridge
pulls every coefficient smoothly toward zero along a hyperbola and
never reaches it, so the full predictor set survives at every penalty;
lasso drives coefficients to exactly zero one after another, so the
model sheds predictors as the penalty rises. The contrast is the whole
practical difference between the two: ridge stabilises an
ill-conditioned fit while keeping every variable, and lasso selects a
sparse subset when many candidates are believed inactive.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-shrinkage-paths.tex}

\paragraph{Worked example: ridge on an orthogonal design, coefficient
by coefficient.} When the predictors are standardised and orthogonal,
so $\mat{X}^{\transpose}\mat{X} = \mat{I}_{p}$, ridge has a
coordinate-wise closed form that makes the shrinkage transparent. The
estimator collapses to $\hat{\beta}_{j}^{\mathrm{ridge}}(\lambda) =
\hat{\beta}_{j}^{\mathrm{OLS}}/(1 + \lambda)$, every coefficient
divided by the same factor. With OLS coefficients $\hat{\vect{\beta}}^{
\mathrm{OLS}} = (2.4, 1.5, 0.8, -1.0)^{\transpose}$ and $\lambda = 1$,
each coefficient halves to $(1.2, 0.75, 0.4, -0.5)^{\transpose}$; at
$\lambda = 4$ each shrinks to a fifth, $(0.48, 0.30, 0.16,
-0.20)^{\transpose}$. The lasso on the same orthogonal design applies
\engterm{soft thresholding} instead, $\hat{\beta}_{j}^{\mathrm{lasso}}
= \operatorname{sgn}(\hat{\beta}_{j}^{\mathrm{OLS}})\max(0,\,
\abs{\hat{\beta}_{j}^{\mathrm{OLS}}} - \lambda/2)$, subtracting a fixed
amount from each coefficient's magnitude and clipping at zero. At
$\lambda/2 = 0.9$ the lasso estimates are $(1.5, 0.6, 0, -0.1)^{
\transpose}$: the third coefficient, whose OLS value $0.8$ fell below
the threshold, is set exactly to zero while the others are pulled in by
the same $0.9$. The two rules expose the geometric difference the
figure draws. Ridge scales, so a coefficient never reaches zero and
the relative ordering of the coefficients is preserved; lasso
translates, so small coefficients vanish and the surviving ones are
biased toward zero by a constant, which is the selection-versus-scaling
distinction the engineer trades on when choosing between them.

The choice of $\lambda$ for either estimator is the central
modelling decision, and cross-validation is the standard discipline
for making it.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-bias-variance.tex}

The ridge penalty was introduced by Hoerl and Kennard for
nonorthogonal design matrices in chemical engineering
\cite{paper:v2ch14-hoerl-kennard-ridge-1970}; the lasso, with
its variable-selection property, was introduced by Tibshirani for
problems where many candidate predictors are believed to be
inactive \cite{paper:v2ch14-tibshirani-lasso-1996}. The two
penalties are now the working catalogue's first reach; a fuller
treatment with elastic-net, group-lasso, and adaptive variants
sits in Hastie, Tibshirani, and Friedman
\cite{text:hastie-tibshirani-friedman}.

\begin{estimation}
\textbf{Question.} A ridge regression with
$\mat{X}^{\transpose}\mat{X}$ approximately $\mat{I}_{p}$ and
$p = 10$ has unbiased OLS variance per coefficient of
$\sigma^{2}/n = 1.0$. Ridge introduces a bias whose square scales
as $b^{2}(\lambda) \approx (\lambda/(1+\lambda))^{2}\,
\norm*{\vect{\beta}}_{2}^{2}/p$ at moderate $\lambda$, with
$\norm*{\vect{\beta}}_{2}^{2}/p \approx 0.5$ in working units.
Estimate the value of $\lambda$ that approximately minimises the
mean squared error per coefficient.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
Variance scales as $V(\lambda) \approx 1/(1+\lambda)^{2}$ in the
orthogonal-design approximation; bias squared scales as
$B^{2}(\lambda) \approx 0.5\,(\lambda/(1+\lambda))^{2}$. Setting
$\dd(V + B^{2})/\dd\lambda \approx 0$ gives $-2/(1+\lambda)^{3} +
0.5 \cdot 2\lambda/(1+\lambda)^{3} = 0$, hence $\lambda^{\star}
\approx 1/0.5 = 2$. The minimum is broad; any $\lambda$ in the
range $1$ to $4$ produces a mean squared error within $10\,\%$ of
the optimum. The lesson the engineer takes home is the shape of
the curve, not the exact minimum: cross-validation will land in
the broad valley provided the procedure is honest, and reporting
a single $\lambda$ without its cross-validated error band is
overspecifying the choice.
\end{estimation}

\section{Cross-validation, hold-out, the danger of looking at test data}\pathtag{core}

The MLE and the OLS estimator both have closed-form measures of
in-sample error: the residual sum of squares, the fitted likelihood,
the in-sample mean squared error. None of these is what the
engineer needs. The engineer needs the error the model will produce
on data it has not seen.

\subsection{Training error versus generalisation error}

The \engterm{training error} of an estimator is the average loss on
the data used to fit the model. The \engterm{generalisation error},
or expected prediction error, is the average loss on a fresh
observation drawn from the same population. The training error
underestimates the generalisation error, often substantially, and
the gap grows with model flexibility \cite[ch.~21]{text:wasserman2004}.

The phenomenon has a working name. A model selected on training
error is selected on its ability to memorise the training data. The
fit's variance shows up as part of the training fit, not as part of
the residual. \engterm{Overfitting} is the technical name for the
mismatch.

\subsection{Hold-out}

The \engterm{hold-out} discipline partitions the data into a
\engterm{training set} and a \engterm{test set}. The model is fit
on the training set, including any tuning of hyperparameters
($\lambda$ in ridge or lasso, the polynomial degree in a polynomial
regression). The fitted model is evaluated once on the test set,
and the test error is reported as the estimate of generalisation
error.

The discipline has one rule. The test set is looked at exactly
once, after every choice has been made. An engineer who tunes a
hyperparameter on the test set, refits, and looks again has
contaminated the procedure. The test error is then biased downward
by an amount that grows with the number of looks. The Volume I Ch
8 ban on choosing the analysis after seeing the data is the same
discipline at the level of model selection.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-kfold.tex}

\subsection{$k$-fold cross-validation}

Hold-out wastes data. \engterm{$k$-fold cross-validation} partitions
the data into $k$ folds of approximately equal size, fits the model
$k$ times (each time holding out one fold and fitting on the union
of the other $k-1$), and averages the held-out errors:
\[
\widehat{\mathrm{CV}}_{k}
 = \frac{1}{k}\sum_{j=1}^{k}
   \frac{1}{\abs*{\mathcal{F}_{j}}}
   \sum_{i \in \mathcal{F}_{j}}
   L\!\left(y_{i},\, \hat{f}^{(-j)}(x_{i})\right),
\]
where $\hat{f}^{(-j)}$ is the model fit on the data with fold
$\mathcal{F}_{j}$ removed and $L$ is the working loss (squared
error for regression, log-loss for classification, misclassification
rate for a hard classifier). The standard choices are $k = 5$ and
$k = 10$ \cite[ch.~21]{text:wasserman2004}. The case $k = n$ is
\engterm{leave-one-out cross-validation}, computationally cheap for
linear models with a closed-form leverage formula and expensive
otherwise.

The cross-validated error is itself an estimator and has its own
standard error. A common rule, the \engterm{one-standard-error
rule}, is to choose the most parsimonious model whose
cross-validated error is within one cross-validation standard
error of the minimum. The rule trades a small increase in expected
error for a substantial reduction in model complexity.

\paragraph{Worked example: a cross-validation table.} A
polynomial-degree selection on $n = 60$ points runs $5$-fold
cross-validation for degrees one through six. The held-out mean
squared error per fold, in working units, comes out as

\begin{center}
\small
\begin{tabular}{lcccccc}
\toprule
degree & fold 1 & fold 2 & fold 3 & fold 4 & fold 5 & CV mean (se) \\
\midrule
1 & 0.81 & 0.92 & 0.78 & 0.88 & 0.85 & $0.848\ (0.024)$ \\
2 & 0.44 & 0.49 & 0.41 & 0.52 & 0.46 & $0.464\ (0.019)$ \\
3 & 0.21 & 0.25 & 0.19 & 0.27 & 0.23 & $0.230\ (0.014)$ \\
4 & 0.22 & 0.28 & 0.20 & 0.30 & 0.24 & $0.248\ (0.018)$ \\
5 & 0.29 & 0.41 & 0.26 & 0.44 & 0.33 & $0.346\ (0.035)$ \\
6 & 0.48 & 0.71 & 0.39 & 0.78 & 0.55 & $0.582\ (0.073)$ \\
\bottomrule
\end{tabular}
\end{center}

The cross-validated error falls from degree one to a minimum at
degree three, then rises as the higher-degree models begin fitting
noise. The minimum is $0.230$ at degree three with cross-validation
standard error $0.014$. The one-standard-error rule asks for the
most parsimonious model whose CV error is within $0.230 + 0.014 =
0.244$ of nothing simpler beats it; degree two at $0.464$ does not
qualify, so the rule selects degree three here. The widening
standard errors at degrees five and six are themselves a signal:
the high-degree fits are unstable across folds, which is the
fingerprint of variance, the quantity cross-validation exists to
expose.

\begin{mastery}
The leave-one-out cross-validation error for a linear smoother has
a closed form that makes the connection between leverage and
generalisation explicit. If $\hat{y}_{i} = \sum_{j} H_{ij} y_{j}$
for a hat matrix $\mat{H}$ that does not depend on $y$, then the
leave-one-out residual equals the ordinary residual divided by one
minus the leverage:
\[
y_{i} - \hat{y}_{i}^{(-i)} = \frac{y_{i} - \hat{y}_{i}}{1 - H_{ii}},
\qquad
\widehat{\mathrm{CV}}_{n} = \frac{1}{n}\sum_{i=1}^{n}
\left(\frac{y_{i} - \hat{y}_{i}}{1 - H_{ii}}\right)^{2}.
\]
The full leave-one-out estimate costs a single fit rather than $n$
refits, and the formula shows why high-leverage points dominate the
generalisation estimate: an observation with $H_{ii}$ near one is
nearly fit by itself, its leave-one-out residual blows up, and the
cross-validation error rightly penalises a model that leans on a
single point. The effective degrees of freedom $\tr(\mat{H})$ is
the sum of the leverages and is the model-complexity quantity that
Mallows' $C_{p}$, the Akaike information criterion, and ridge's
$\tr(\mat{H}) = \sum_{j}\sigma_{j}^{2}/(\sigma_{j}^{2} + \lambda)$
all measure. Cross-validation estimates the same generalisation
target without committing to a noise-variance estimate, which is its
advantage when $\sigma^{2}$ is itself uncertain.
\end{mastery}

\subsection{Data leakage}

\engterm{Data leakage} is the inadvertent flow of information from
the test set into the model fit. The canonical leakage modes are:

\begin{itemize}[leftmargin=*]
  \item Pre-processing fitted on the full dataset (mean-centring,
    scaling, principal-component projections) before the
    train-test split.
  \item Imputation of missing values using statistics computed on
    the full dataset.
  \item Variable selection on the full dataset, followed by
    cross-validation on the selected variables. The selection
    itself contaminated the procedure; the cross-validation is
    optimistic.
  \item Tuning hyperparameters on the test set. The test error is
    no longer an estimate of generalisation; it is an estimate of
    how well the model fits the data the engineer has already
    optimised against.
  \item Time-series cross-validation that mixes past and future,
    in datasets where the temporal order carries information.
\end{itemize}

The discipline is direct. Every operation that touches the response
or that adapts to the data must sit inside the cross-validation
loop. A pre-processing step that sees only the training fold within
each iteration is the working pattern. A pre-processing step
applied once to the whole dataset before splitting is leakage.

\begin{principle}[Cross-validation is a procedure, not a number]
The cross-validated error is reportable only when the procedure
that produced it touched no part of the test data, fold by fold.
Every choice that adapts to the data, including pre-processing and
feature selection, must sit inside the loop. The engineer who
reports a cross-validated error from a procedure that violated
this rule is reporting an estimate of training error with a
confidence-inflating disguise.
\end{principle}

\begin{mastery}
Cross-validation estimates an average error; it does not by itself
attach a valid interval to a single new prediction. \engterm{Conformal
prediction} supplies one, and it does so without any distributional
assumption, which is what distinguishes it from the parametric
prediction interval of section 14.1. The split-conformal recipe is
short. Partition the data into a training set and a held-out
\engterm{calibration set}. Fit the model on the training set, and on
each of the $m$ calibration points compute a \engterm{nonconformity
score}, for regression the absolute residual $r_{j} = \abs{y_{j} -
\hat{f}(x_{j})}$. For a target coverage $1 - \alpha$, take the
$\lceil (m+1)(1-\alpha) \rceil / m$ empirical quantile $\hat{q}$ of the
calibration scores. The prediction interval for a new point is then
$\hat{f}(x_{\mathrm{new}}) \pm \hat{q}$, and its finite-sample coverage
is at least $1 - \alpha$ under the single assumption that the
calibration points and the new point are exchangeable, no Gaussian
tail, no correct model, no asymptotics required. The guarantee is
\engterm{marginal}, averaged over the randomness in the calibration
draw, and the interval has constant width unless the score is
normalised by a local spread estimate to make it adaptive. The
mechanism is the same rank argument the permutation test used: the new
point's score is equally likely to fall at any rank among the $m + 1$
scores under exchangeability, so the probability it exceeds the
$\lceil (m+1)(1-\alpha)\rceil$-th is at most $\alpha$ by counting
alone. Conformal prediction is the working tool when the engineer needs
a prediction interval with a coverage guarantee on a model whose error
distribution is unknown or plainly non-Gaussian, a black-box regressor
or a machine-learned classifier, where the parametric prediction
interval has no license and the bootstrap prediction interval carries
only an asymptotic one. Its limitation is the exchangeability
assumption itself: a distribution shift between calibration and
deployment, or a time-ordered series where the future is not
exchangeable with the past, breaks the guarantee exactly as it breaks
every other resampling method in this chapter.
\end{mastery}

\section{A worked case study: from data to a defensible interval}\pathtag{standard}

The apparatus of the chapter earns its keep on a single problem
carried from raw data to a reportable conclusion with every
assumption checked along the way. An engineer is asked the working
question: at a stress amplitude of $260\,\si{\mega\pascal}$, what
fatigue life can the design depend on for a new aluminium alloy, and
with what uncertainty? The dataset is twenty constant-amplitude
fatigue tests, each recording the applied stress amplitude $S$ (in
megapascals) and the number of cycles to failure $N$. The data are
stored at
\repopath{volumes/02-form/ch14-statistical-inference/data/fatigue-sn.csv}
and the full analysis script at
\repopath{volumes/02-form/ch14-statistical-inference/code/casestudy\_fatigue.py}.

\subsection{Step one: state the model before looking at the residuals}

Fatigue life under constant-amplitude loading follows, over the
high-cycle regime, an approximately power-law relation between stress
and life, the Basquin form $N = c\,S^{-m}$. Taking base-ten
logarithms linearises it:
\[
\log_{10} N = \beta_{0} + \beta_{1}\log_{10} S + \varepsilon,
\qquad \beta_{1} = -m.
\]
The transform is chosen before the fit and for two reasons that the
engineer states in advance. The physical model is multiplicative, so
the additive-Gaussian-error assumption is far more defensible on the
log scale than on the raw cycle count, where a test at $10^{5}$
cycles and a test at $10^{3}$ cycles cannot share a common error
variance. Naming the transform before seeing the residuals is the
discipline that keeps the analysis out of the garden of forking
paths of section 14.7: the choice is justified by the mechanism, not
selected after the residual plot rewarded it.

\subsection{Step two: fit and read the parametric uncertainty}

The ordinary least-squares fit on the twenty $(\log_{10} S,
\log_{10} N)$ pairs returns $\hat{\beta}_{0} = 19.84$ and
$\hat{\beta}_{1} = -6.20$, so the fitted Basquin exponent is
$\hat{m} = 6.20$, within the range tabled for wrought aluminium
alloys. The residual standard deviation on the log scale is
$\hat{\sigma} = 0.115$, the centred predictor sum of squares is
$S_{xx} = 0.100$ in $(\log_{10} S)^{2}$ units, and the standard
error of the slope is $\hat{\sigma}/\sqrt{S_{xx}} = 0.115/0.317 =
0.362$. A $95\,\%$ confidence interval for the slope, with
$t_{0.975,18} = 2.101$, is $-6.20 \pm 2.101 \cdot 0.362 = [-6.96,
-5.44]$, an interval on the Basquin exponent that the materials
engineer can compare against handbook values. The coefficient of
determination is $R^{2} = 0.94$, high enough to invite the
overconfidence the next step guards against.

\subsection{Step three: check the assumptions the fit rests on}

A high $R^{2}$ is not a license. The fit assumes the log-scale
errors are homoscedastic, Gaussian, and independent, and each
assumption is checked, not assumed. Figure
\ref{fig:vol02:ch14:casestudy-diagnostics} carries the two plots
that do most of the work. The residuals-against-fitted panel shows a
band of roughly constant width about zero with no funnel and no
curve, so the homoscedasticity and the linearity of the log-log
relation both survive. The normal quantile-quantile panel tracks the
reference line through the bulk of the range with a single
upper-tail point above the line, a mild departure that says the
log-scale error distribution has a slightly heavier upper tail than
Gaussian. Independence is checked separately: the tests were run in a
randomised order across two test frames, and a plot of residuals
against run order (in the script, not reproduced here) shows no
trend, so no run-order or machine drift contaminates the fit. The
upper-tail departure is the finding the engineer carries forward
rather than suppresses, because it is the reason the parametric
prediction interval, which leans on Gaussian tails, is checked
against a distribution-free alternative in the next step.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-casestudy-diagnostics.tex}

\subsection{Step four: the prediction, with the right interval}

The design question asks for the life at $S = 260\,\si{\mega\pascal}$,
so $\log_{10} S = 2.415$. The point prediction is $\widehat{\log_{10}
N} = 19.84 - 6.20 \cdot 2.415 = 4.88$, that is $N \approx 7.6 \times
10^{4}$ cycles. The design needs a \emph{prediction} interval for a
single future specimen, not a confidence interval for the mean life,
because a part in service is one draw, not an average. The prediction
standard error at the query point is
\[
\widehat{\mathrm{se}}_{\mathrm{pred}} = \hat{\sigma}\sqrt{1 +
\frac{1}{n} + \frac{(\log_{10} S - \overline{\log_{10} S})^{2}}
{S_{xx}}}
= 0.115\sqrt{1 + 0.05 + 0.001} = 0.115 \cdot 1.025 = 0.118,
\]
where the query stress sits near the centre of the tested range
($\overline{\log_{10} S} = 2.424$), so the leverage term is small.
The $95\,\%$ parametric prediction interval on the log scale is
$4.88 \pm 2.101 \cdot 0.118 = [4.63, 5.13]$, which back-transforms
to a life interval of $[4.3 \times 10^{4},\, 1.3 \times 10^{5}]$
cycles. Because the quantile-quantile
plot warned of a heavier upper tail, the engineer recomputes the
interval by the bootstrap: resampling the twenty residuals with
replacement, adding them to the fitted values, refitting, and reading
the $2.5$th and $97.5$th percentiles of the resulting predicted-life
distribution gives an interval close to $[4.7 \times 10^{4},\, 1.3
\times 10^{5}]$ cycles, narrower than the parametric interval on the
low side because the bootstrap does not impose the Gaussian lower
tail. The two intervals agree to within the precision the design
needs, which is itself the evidence that the mild non-normality does
not overturn the parametric result; the engineer reports the
parametric interval as the headline and notes that the
distribution-free check confirmed it.

\subsection{Step five: the uncertainty statement, written correctly}

The deliverable is a sentence the engineer can defend. At a stress
amplitude of $260\,\si{\mega\pascal}$, the fitted model predicts a
median fatigue life near $7.6 \times 10^{4}$ cycles, with a $95\,\%$
prediction interval for a single specimen of roughly $4 \times
10^{4}$ to $1.4 \times 10^{5}$ cycles, valid over the tested stress
range of $200$ to $340\,\si{\mega\pascal}$ for this alloy and surface
finish, and not extrapolable below the lowest tested amplitude where
the run-out regime begins. The statement names the quantity (a
single-specimen life, not a mean), the level ($95\,\%$), the
interval, the domain of validity, and the explicit boundary beyond
which the model does not speak. A design that needs a life floor uses
the lower bound, $4 \times 10^{4}$ cycles, with whatever additional
safety factor the consequence of failure demands, and never the point
estimate alone. The contrast with the naive report, ``$R^{2} = 0.96$,
predicted life $7.4 \times 10^{4}$ cycles,'' is the entire lesson of
the chapter compressed into one comparison: the naive report omits
the interval, the interval's kind, the domain, and the boundary, and
each omission is a place a part fails earlier than the report
promised.

\section{A second case study: a censored time-to-failure analysis}\pathtag{standard}

The fatigue study fitted a regression to complete data. The
reliability engineer rarely has complete data, because a life test
ends before every unit has failed, and the units still running carry
\engterm{right-censored} observations: their failure time is known
only to exceed the test duration. The censored observation is real
information, not a missing value, and discarding it biases every life
estimate downward. This second case carries a censored life test from
raw data to a confidence-bounded life estimate, with the censoring
handled correctly at every step, and it exercises the
maximum-likelihood, delta-method, and bootstrap machinery on a
problem whose structure the regression case did not reach. The data
are stored at
\repopath{volumes/02-form/ch14-statistical-inference/data/bearing-life.csv}
and the analysis script at
\repopath{volumes/02-form/ch14-statistical-inference/code/casestudy\_weibull.py}.

\subsection{Step one: the model and the censored likelihood}

A batch of $n = 10$ ball bearings is run on a test rig to a planned
cutoff of $500$ hours. Eight bearings fail at recorded times, in
hours, $\set{120, 180, 226, 270, 314, 358, 408, 484}$, and two are
still running at the $500$-hour cutoff, contributing two
right-censored observations at $500$. Bearing life under rolling
contact fatigue is conventionally modelled by the \engterm{Weibull
distribution} \cite{paper:weibull1939}, whose flexibility in the
shape of the hazard makes it the reliability engineer's default. The
Weibull has shape $\beta$ and scale (characteristic life) $\eta$, with
density, survival function, and hazard
\[
f(t) = \frac{\beta}{\eta}\!\left(\frac{t}{\eta}\right)^{\beta-1}
e^{-(t/\eta)^{\beta}},
\quad
S(t) = e^{-(t/\eta)^{\beta}},
\quad
h(t) = \frac{\beta}{\eta}\!\left(\frac{t}{\eta}\right)^{\beta-1}.
\]
The shape carries the engineering meaning: $\beta < 1$ is a falling
hazard (infant mortality), $\beta = 1$ is the constant-hazard
exponential (random failures), and $\beta > 1$ is a rising hazard
(wear-out), which is what a bearing on a fatigue rig should show. The
likelihood splits the two kinds of observation. A failure at time
$t_{i}$ contributes its density $f(t_{i})$; a unit censored at time
$c_{j}$ contributes its survival probability $S(c_{j})$, the
probability that it lasted at least that long. With $d$ failures and
the rest censored,
\[
L(\beta, \eta) = \prod_{i \in \text{failed}} f(t_{i})
                 \prod_{j \in \text{censored}} S(c_{j}),
\]
and the log-likelihood is maximised numerically, since no closed form
exists once censoring enters. Naming the censored term before the fit
is the discipline: an analyst who drops the two survivors and fits the
eight failures alone is fitting the wrong likelihood and will report a
characteristic life biased low.

\subsection{Step two: a graphical check before the numerical fit}

The Weibull probability plot transforms the axes so that a Weibull
sample falls on a straight line. Taking the logarithm of the survival
function twice gives $\ln(-\ln S(t)) = \beta\ln t - \beta\ln\eta$, a
line in $\ln t$ with slope $\beta$ and intercept $-\beta\ln\eta$. The
engineer assigns each ordered failure a plotting position by the
median-rank formula $\hat{F}_{i} = (i - 0.3)/(n + 0.4)$ on the full
sample size $n = 10$, so the two survivors keep the late failures'
plotting positions from being pushed to one, and plots
$\ln(-\ln(1 - \hat{F}_{i}))$ against
$\ln t_{i}$. Figure \ref{fig:vol02:ch14:weibull-plot} shows the eight
points falling close to a line, the graphical licence to trust the
single-Weibull model. A systematic curve, a dogleg with two slopes,
would have been the signature of two competing failure modes and would
have sent the engineer to a mixture model before any single-Weibull
number was reported.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-weibull-plot.tex}

The Weibull plot presumes a Weibull. A stricter check makes no
distributional assumption at all, and the engineer runs it before
committing to the parametric family. The \engterm{Kaplan-Meier}
estimator, or product-limit estimator, reconstructs the survival
function $S(t) = \Pr[T > t]$ from censored data using only the order of
the failures and the count still at risk. At each failure time $t_{(i)}$
with $d_{i}$ failures among $n_{i}$ units still at risk just before it,
the estimator multiplies the survival by the factor $(1 - d_{i}/n_{i})$,
\[
\hat{S}(t) = \prod_{t_{(i)} \leq t}\left(1 - \frac{d_{i}}{n_{i}}\right),
\]
and holds the step flat between failures, censored units leaving the
at-risk count without triggering a drop. For the ten bearings, all
distinct single failures, the at-risk count falls from ten by one at
each failure time, so the survival steps down through $0.9, 0.8, \ldots,
0.2$ across the eight failures and holds at $0.2$ past the last failure
because the two units censored at $500$ hours never fail. Figure
\ref{fig:vol02ch14:kaplan-meier} overlays the fitted Weibull survival
curve on this step, and the two tracking each other through the
observed range is the distribution-free licence for the parametric
model that the Weibull plot only suggested. A parametric curve that
pulled away from the Kaplan-Meier step, sitting above it in the middle
and below it in the tail, would be the signal that the single-Weibull
shape is wrong, sending the engineer to a two-parameter mixture or a
different family before any characteristic-life number is quoted. The
nonparametric estimate has its own uncertainty, the
\engterm{Greenwood} variance $\widehat{\Var}[\hat{S}(t)] = \hat{S}(t)^{2}
\sum_{t_{(i)} \leq t} d_{i}/(n_{i}(n_{i} - d_{i}))$, which widens the
pointwise band toward the tail where few units remain at risk, and it is
the reason a Kaplan-Meier curve near its right end is read with the same
caution the Weibull fit's lower-tail extrapolation earns.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-kaplan-meier.tex}

\subsection{Step three: the maximum-likelihood fit and its information}

Maximising the censored log-likelihood numerically returns
$\hat{\beta} = 2.33$ and $\hat{\eta} = 404$ hours. The shape above one
confirms the rising, wear-out hazard the bearing physics predicts. The
standard errors come from the observed information, the negative
Hessian of the log-likelihood at the maximum, inverted to a $2\times2$
covariance matrix; for this dataset the routine returns
$\widehat{\mathrm{se}}(\hat{\beta}) = 0.70$ and
$\widehat{\mathrm{se}}(\hat{\eta}) = 63$ hours, with a mild negative
correlation between the two estimates. A Wald interval for the shape
is $2.33 \pm 1.96 \cdot 0.70 = [0.96, 3.70]$, which the engineer reads
with caution: its lower end nearly reaches one, but the symmetric Wald
approximation strains on a small censored sample, and the
likelihood-ratio interval, found by profiling the log-likelihood, is
the one to trust here because it respects the parameter's boundary at
$\beta > 0$ and the skew of its sampling distribution.

\subsection{Step four: the quantity the design needs, with bounds}

A bearing-life specification rarely asks for the characteristic life.
It asks for a low percentile, the \engterm{$B_{10}$ life}, the time by
which ten percent of the population has failed, because the design
must protect the early failures, not the median. For the Weibull,
$t_{p} = \eta\,(-\ln(1 - p))^{1/\beta}$, so the $B_{10}$ life is
\[
\hat{t}_{0.10} = \hat{\eta}\,(-\ln 0.90)^{1/\hat{\beta}}
= 404 \cdot (0.10536)^{1/2.33}
= 404 \cdot 0.381 = 154 \text{ hours}.
\]
The point estimate is not the deliverable; the lower confidence bound
is, because a design that depends on the $B_{10}$ life must protect
against the estimate being optimistic. The $B_{10}$ life is a
nonlinear function of both fitted parameters, so its standard error
follows from the delta method applied to $g(\beta, \eta) = \eta\,
(-\ln 0.9)^{1/\beta}$, propagating the full $2\times2$ covariance
through the gradient $\nabla g$. The delta-method calculation returns
a standard error near $52$ hours, giving an approximate $95\,\%$
lower confidence bound of $154 - 1.645 \cdot 52 \approx 68$ hours for
a one-sided design limit. Because the $B_{10}$ life sits in the lower
tail where the Weibull sampling distribution is skewed, the engineer
cross-checks the delta-method bound against a parametric bootstrap:
simulate many censored datasets from the fitted Weibull, refit each,
and read the $5$th percentile of the resulting $B_{10}$ estimates,
which lands near $88$ hours, above the delta-method bound because the
bootstrap captures the lower-tail skew the symmetric delta interval
overcorrects. The engineer reports the more conservative
delta-method bound of $68$ hours as the design floor and notes that
the bootstrap places the bound no lower, which is the cross-check
discipline the fatigue case applied to its prediction interval.

\subsection{Step five: the defensible sentence}

The deliverable reads: on a Weibull model fitted to ten bearings with
two right-censored at the $500$-hour cutoff, the fitted shape is
$\hat{\beta} = 2.33$, consistent with wear-out, the characteristic
life is $\hat{\eta} = 404$ hours, and the $B_{10}$ life is estimated
at $154$ hours with a one-sided $95\,\%$ lower confidence bound near
$68$ hours, valid for this bearing under the test rig's load and
speed and not extrapolable to field conditions that differ in
lubrication or contamination. The sentence names the model, the
censoring, the shape (which carries the failure-mode claim), the
design-relevant percentile rather than the median, the confidence
bound rather than the point estimate, and the boundary of validity.
The naive report, ``mean time to failure $295$ hours'' computed by
averaging the eight failure times and ignoring the two survivors,
commits two errors at once: it discards the censored information, and
it reports a central tendency where the design needs a protected lower
percentile. The gap between $295$ hours of naive mean and $68$ hours
of defensible $B_{10}$ lower bound is the distance between a number
and an engineering decision.

\section{Designed experiments and response-surface optimisation}\pathtag{standard}

The two-way analysis of variance read the effect of factors that were
set deliberately. The full discipline of choosing those settings to
extract the most information per run is \engterm{design of
experiments}, and its optimisation arm, \engterm{response-surface
methodology}, is the engineer's tool for finding the process settings
that maximise a response \cite[ch.~11]{text:montgomery-doe-2019}. The
method earns a worked example because it closes the loop from
estimation to action: the regression fits a surface, and the fitted
surface points to where to run next.

\subsection{The two-level factorial and its efficiency}

A \engterm{two-level full factorial} runs every combination of $k$
factors each at a low and a high setting, $2^{k}$ runs in all. With
two factors the four corner runs estimate the grand mean, both main
effects, and the interaction, each from the full set of runs rather
than from a one-factor-at-a-time sweep that would waste runs and miss
the interaction entirely. The factorial's efficiency is the reason the
design literature treats one-factor-at-a-time experimentation as a
mistake: changing one factor while holding the others fixed estimates
each effect from half the data and cannot see whether the factors
interact. When the factor count is large, a \engterm{fractional
factorial} runs a carefully chosen subset, $2^{k-p}$ runs, accepting
that some high-order interactions are \engterm{aliased} (confounded)
with lower-order effects, a trade the engineer makes deliberately by
choosing the design's resolution.

\subsection{Steepest ascent and the central composite design}

A factorial fits a first-order plane, $\hat{y} = \hat{\beta}_{0} +
\hat{\beta}_{1}x_{1} + \hat{\beta}_{2}x_{2}$, whose gradient points
uphill. The \engterm{method of steepest ascent} moves the experiment
in the gradient direction until the response stops improving, which is
the signal that the linear approximation has reached the neighbourhood
of a maximum and curvature now matters. To fit curvature the engineer
augments the factorial with centre points and \engterm{axial points},
producing a \engterm{central composite design} that supports a full
second-order model,
\[
\hat{y} = \hat{\beta}_{0} + \sum_{j}\hat{\beta}_{j}x_{j}
+ \sum_{j}\hat{\beta}_{jj}x_{j}^{2}
+ \sum_{j<l}\hat{\beta}_{jl}x_{j}x_{l}.
\]
The centre points, replicated, supply a pure-error estimate and a
test for curvature; the axial points supply the spread along each axis
needed to estimate the quadratic terms.

\paragraph{Worked example: maximising a reaction yield.} A process
yield depends on temperature and reaction time, coded so that the
factorial corners sit at $\pm 1$. A central composite design fits the
second-order surface
\[
\hat{y} = 91.0 + 1.8\,x_{1} - 2.4\,x_{2} - 3.0\,x_{1}^{2}
- 2.0\,x_{2}^{2} - 1.5\,x_{1}x_{2},
\]
where $y$ is the yield in percent and $x_{1}, x_{2}$ are coded
temperature and time. The stationary point, where both partial
derivatives vanish, solves the linear system
\[
\pdv{\hat{y}}{x_{1}} = 1.8 - 6.0\,x_{1} - 1.5\,x_{2} = 0,
\qquad
\pdv{\hat{y}}{x_{2}} = -2.4 - 1.5\,x_{1} - 4.0\,x_{2} = 0.
\]
Solving, from the first equation $x_{1} = (1.8 - 1.5\,x_{2})/6.0$;
substituting into the second gives $-2.4 - 1.5(1.8 - 1.5\,x_{2})/6.0
- 4.0\,x_{2} = 0$, that is $-2.4 - 0.45 + 0.375\,x_{2} - 4.0\,x_{2}
= 0$, so $-2.85 = 3.625\,x_{2}$ and $x_{2} = -0.79$; back-substituting
gives $x_{1} = (1.8 + 1.185)/6.0 = 0.50$. The stationary point sits at
coded $(0.50, -0.79)$, inside the experimental region. To confirm it is
a maximum rather than a saddle, the engineer inspects the matrix of
second-order coefficients
\[
\mat{B} = \begin{pmatrix} \hat{\beta}_{11} & \tfrac{1}{2}\hat{\beta}_{12}
\\ \tfrac{1}{2}\hat{\beta}_{12} & \hat{\beta}_{22} \end{pmatrix}
= \begin{pmatrix} -3.0 & -0.75 \\ -0.75 & -2.0 \end{pmatrix},
\]
whose eigenvalues are both negative ($-3.40$ and $-1.60$), so the
surface is concave and the stationary point is a maximum. The
predicted yield there is $\hat{y} \approx 92.4\,\%$, against the
$91.0\,\%$ at the design centre, and the eigenvector of the smaller
eigenvalue marks the flat direction along which the process tolerates
drift. Figure \ref{fig:vol02:ch14:response-surface} draws the fitted
contours; the engineer reads the optimum's location and the contour
spacing together, because a tight contour set means the yield is
sensitive to a setting error the operator might make, and a flat
optimum means it is not. The next experiment is run at the predicted
optimum to confirm it, never reported as established from the fitted
surface alone, since the surface is an extrapolating quadratic and the
confirmation run is the cheap insurance against a fitted maximum that
the process does not actually deliver.

\input{volumes/02-form/ch14-statistical-inference/figures/fig-response-surface.tex}

\begin{mastery}
The percentile bootstrap interval that section 14.1 introduced is the
crudest of the bootstrap intervals, and two refinements matter when
the statistic is biased or skewed. The \engterm{bias-corrected and
accelerated} interval, the \engterm{BCa} interval, adjusts the
percentiles it reads off the bootstrap distribution by two quantities
estimated from the resamples \cite{paper:ch08-efron-bca-1987}. The
bias-correction $\hat{z}_{0}$ is the normal quantile of the fraction
of bootstrap replicates below the original estimate, zero when the
bootstrap distribution is centred on the estimate and nonzero when the
statistic is biased. The acceleration $\hat{a}$ measures how fast the
standard error changes with the parameter, estimated from the
\engterm{jackknife} skewness of the statistic. The jackknife itself is
the leave-one-out cousin of the bootstrap: it recomputes the statistic
on each of the $n$ subsamples that omit one observation, and the
spread of those $n$ values estimates the standard error, with the
jackknife's third moment supplying the acceleration BCa needs. The BCa
interval is transformation-invariant and second-order accurate, where
the percentile interval is only first-order accurate, so the
refinement is worth the extra computation precisely when the
percentile and BCa intervals disagree, which is the diagnostic that
the statistic's sampling distribution is skewed enough to matter. A
deeper variant, the \engterm{bootstrap-t}, studentises each replicate
by its own standard-error estimate and is more accurate still for a
location statistic, at the cost of a nested bootstrap to estimate the
per-replicate standard error. The working rule is to default to BCa
for a confidence interval on a biased or bounded statistic, to fall
back to the percentile interval only when the statistic is plainly
symmetric, and to read a large gap between the two as evidence that
the simple interval was hiding the skew.
\end{mastery}

\section{Failure: p-hacking, multiple comparisons, replication crisis revisited}\pathtag{core}

Volume I Chapter 8 named four malpractices: p-hacking, the garden
of forking paths, HARKing, and inappropriate stopping rules. The
mathematical apparatus of this chapter lets us state each one
formally, attach the controlling correction, and connect the
catalogue to the replication studies that have measured the size of
the problem in published literatures.

\subsection{P-hacking, formally}

Let $T_{1}, \ldots, T_{m}$ be a family of test statistics computed
on the same dataset. \engterm{P-hacking} is the practice of
selecting the smallest of $m$ associated $p$-values and reporting
it as a single test result \cite{method:simmons-nelson-simonsohn-2011}.
Under the global null and approximate independence of the tests,
the probability that at least one of $m$ p-values is below $\alpha$
is approximately $1 - (1-\alpha)^{m}$, which exceeds $\alpha$ for
$m > 1$ and approaches one as $m$ grows. Reporting one selected
$p$-value out of $m$ without disclosing $m$ misrepresents the size
of the test.

The Simmons, Nelson, and Simonsohn 2011 simulation study quantified
the inflation: under realistic combinations of optional reporting
of conditions, optional inclusion or exclusion of covariates,
optional sample-size augmentation, and optional choice of dependent
variable, the false-positive rate of a nominally $\alpha = 0.05$
analysis can rise above $60\,\%$
\cite{method:simmons-nelson-simonsohn-2011}.

\subsection{Multiple comparisons and the Bonferroni correction}

When the engineer genuinely needs to test $m$ hypotheses, the
\engterm{family-wise error rate} is the probability of at least one
false rejection. The \engterm{Bonferroni correction} controls the
family-wise error rate at level $\alpha$ by testing each hypothesis
at level $\alpha/m$. The correction is conservative, especially
under positive dependence among the tests, but it is a defensible
default \cite[ch.~10]{text:wasserman2004}.

A more efficient discipline is \engterm{false discovery rate}
control. The Benjamini-Hochberg procedure orders the $m$ p-values
$p_{(1)} \leq \cdots \leq p_{(m)}$, finds the largest $k$ such that
$p_{(k)} \leq k\alpha/m$, and rejects the corresponding $k$
hypotheses. The expected proportion of false rejections among the
rejections is then bounded by $\alpha$. The procedure is the working
standard in genomics, neuroscience, and other fields with thousands
of simultaneous tests.

\paragraph{Worked example: Bonferroni against Benjamini-Hochberg.}
Ten tests return the ordered $p$-values
\[
0.001,\ 0.008,\ 0.011,\ 0.019,\ 0.022,\ 0.031,\ 0.045,\ 0.061,\
0.220,\ 0.530.
\]
At family-wise level $\alpha = 0.05$ with $m = 10$, the Bonferroni
threshold is $0.005$, so only the first $p$-value ($0.001$) is
rejected; the procedure controls the family-wise error rate but
rejects little. The Benjamini-Hochberg procedure compares the
$k$-th ordered $p$-value to $k\alpha/m = 0.005k$:
\[
\begin{array}{c|cccccc}
k & 1 & 2 & 3 & 4 & 5 & 6 \\ \hline
p_{(k)} & 0.001 & 0.008 & 0.011 & 0.019 & 0.022 & 0.031 \\
0.005k & 0.005 & 0.010 & 0.015 & 0.020 & 0.025 & 0.030 \\
p_{(k)} \le 0.005k? & \text{y} & \text{y} & \text{y} & \text{y} & \text{y} & \text{n}
\end{array}
\]
The largest $k$ with $p_{(k)} \leq 0.005k$ is $k = 5$ (at $k = 6$,
$0.031 > 0.030$), so Benjamini-Hochberg rejects the first five
hypotheses, bounding the expected false-discovery proportion among
those five at $0.05$. The two procedures answer different
questions: Bonferroni guards against even one false rejection and
accepts low power; Benjamini-Hochberg tolerates a controlled
fraction of false rejections to recover power. The engineer's
choice follows the cost of a single false discovery, not a
convention.

\begin{mastery}
Bonferroni is the crudest family-wise control, and two refinements
recover power without giving up the guarantee. The \engterm{Holm
step-down} procedure orders the $p$-values $p_{(1)} \leq \cdots \leq
p_{(m)}$ and compares $p_{(k)}$ against $\alpha/(m - k + 1)$,
rejecting in sequence until the first failure and accepting all
remaining hypotheses. The first comparison, $p_{(1)}$ against
$\alpha/m$, is the Bonferroni threshold, but every later comparison
uses a larger threshold, so Holm rejects at least as much as
Bonferroni while controlling the same family-wise error rate under
no assumptions on the dependence structure. On the ten ordered
$p$-values of the worked example, Holm compares $0.001$ against
$0.05/10 = 0.005$ (reject), $0.008$ against $0.05/9 = 0.00556$
(fail), and stops, rejecting one hypothesis, the same as Bonferroni
here because the second $p$-value is large; on a list with several
small $p$-values Holm rejects strictly more. The \engterm{Benjamini-
Hochberg} procedure of the main text controls a different and
weaker target, the false discovery rate rather than the family-wise
error rate, which is why it rejects five hypotheses on the same data
\cite{method:ch08-benjamini-hochberg-1995}. The engineer chooses the
target first, family-wise error when any false positive is costly,
false discovery rate when a controlled fraction of false positives
among many discoveries is acceptable, and only then chooses the
procedure that controls it with the most power.
\end{mastery}

\subsection{The garden of forking paths}

\engterm{The garden of forking paths} \cite{method:gelman-loken-forking-paths}
is the analyst's version of the same problem. The analyst does not
run all $m$ tests; the analyst runs one test, but the choice of
test was made after the data were seen, and the choice would have
been different if the data had been different. The reported
$p$-value is conditioned on the actual data, which the analyst
selected the test from. The reference distribution for the test
statistic is the distribution under the null and the analyst's
data-dependent choice rule, not under the null and the test the
analyst happened to run.

The forking-paths argument applies even when no other test was
explicitly performed. The \emph{counterfactual} flexibility, the
set of analyses the engineer would have run if the data had been
different, is what inflates the false-positive rate.

\subsection{HARKing}

\engterm{HARKing}, hypothesising after the results are known, is
the practice of selecting the hypothesis that fits the result and
presenting it as the pre-registered hypothesis
\cite{method:kerr-harking-1998}. HARKing inflates the false-positive
rate in the same way as p-hacking, with the additional cost that
the literature is then populated by hypotheses that look stronger
than the data warrant.

\subsection{The replication studies}

The 2015 \engterm{Reproducibility Project: Psychology} attempted
replications of $100$ studies from three psychology journals. Of
$97$ original studies that had reported a statistically significant
effect, $36\,\%$ produced a statistically significant effect in
replication, and the mean replication effect size was approximately
half the original \cite{method:open-science-collaboration-2015}.

The 2021 \engterm{Reproducibility Project: Cancer Biology}
attempted to replicate $50$ experiments from $23$ high-profile
cancer-biology papers. Replication effects were on average smaller
than original effects by a factor of about three, with mixed
success rates across multiple replication criteria
\cite{method:errington-cancer-biology-2021}.

The pattern extends beyond psychology. The 2016 Camerer et al.\
replication of $18$ laboratory experiments in economics found
that $61\,\%$ of the original results replicated as significant,
with replication effect sizes averaging about two-thirds of the
originals \cite{method:v2ch14-camerer-economics-2016}. The
\textcite{method:ch08-ioannidis-2005} essay had argued, on the
basis of the power and prior-probability arithmetic that Volume I
Chapter 8 already worked through, that most published research
findings in fields with low statistical power and high analytic
flexibility were likely false. The replication studies above are
the post-hoc empirical evidence behind that argument. The Nuzzo
2014 \textit{Nature} feature surveyed the public arc of the
discussion and the working compromises the discipline arrived at
\cite{method:v2ch14-nuzzo-scientific-method-2014}; the 2016 ASA
Statement on $p$-values \cite{method:ch08-wasserstein-asa-2016}
remains the discipline's most-cited formal position.

The four studies, taken together, are the empirical evidence
behind the engineer's working caution. The named statistical
malpractices are not theoretical risks. They have measurable
effects on the literatures the engineer might draw on for design
input.

\subsection{What an engineer does about it}

The chapter's working discipline closes with a small list, applied
operationally:

\begin{itemize}[leftmargin=*]
  \item Pre-register the analysis, including the test, the model,
    the inclusion criteria, and the stopping rule, before the data
    are seen. Record deviations.
  \item Disclose all tests run, not only those that crossed
    threshold. The number $m$ is not separable from the meaning of
    any one $p$-value.
  \item Apply a multiple-comparisons correction whenever multiple
    tests are reported. Bonferroni for small $m$ and high cost of
    false discovery; Benjamini-Hochberg for large $m$ and tolerance
    for some false discoveries among many true ones.
  \item Report effect sizes with confidence intervals. A
    significant $p$-value with a small effect size is rarely a
    load-bearing result.
  \item Treat a single study as evidence, not as decision. Ask
    for replications, prefer cross-validated error to in-sample
    error, and size the engineering decision against a conservative
    defensible effect size.
\end{itemize}

\begin{principle}[The disclosure rule]
A statistical claim is reportable when the analysis that produced
it was specified before the data were seen, when every test run on
the data has been disclosed, when a multiple-comparisons correction
has been applied where appropriate, and when the effect size has
been reported with a confidence interval. A claim that fails any
of these is a working hypothesis to be replicated, not a result to
be acted on.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Model fit with $k$-fold cross-validated error]
The reader fits a regression model (linear, logistic, or
regularised) to a real dataset of their choice, evaluates it with
$k$-fold cross-validation, and produces an honest report. The
project closes Volume II's statistical-inference arc and supplies
the deliverable that earns the reader admission to chapters that
will assume model-fitting fluency.

\textbf{Track}. Simulation with analysis. \textbf{Hazard class}: none.
\textbf{Tools}. Paper, pencil, and a programming environment with
regression primitives: Python with statsmodels or scikit-learn, R,
Julia with GLM.jl, MATLAB, or equivalent.

\textbf{Dataset choice}. The reader chooses a dataset of $n \geq
50$ observations with at least $p = 3$ candidate predictors.
Acceptable sources include the reader's own Volume I project data,
a public dataset from a recognised repository (UCI Machine Learning
Repository, OpenML, a published-paper supplementary file, a
governmental open-data portal), or a clean export from the reader's
working environment. The reader records the dataset, the source,
the date range, and the operational meaning of the response and
each predictor.

\textbf{Model choice}. The reader fits one of the following:
\begin{enumerate}[leftmargin=*]
  \item A linear regression of a continuous response on the
    predictors. Report OLS coefficients, residual diagnostics, and
    the residual standard deviation.
  \item A logistic regression of a binary response on the
    predictors. Report the coefficients on the log-odds scale, a
    confusion matrix at a stated decision threshold, and the
    classification accuracy.
  \item A ridge or lasso regression of a continuous response on
    the predictors, with the regularisation parameter $\lambda$
    chosen by cross-validation. Report the chosen $\lambda$, the
    coefficient profile as $\lambda$ varies, and the variables
    retained at the chosen $\lambda$.
\end{enumerate}

\textbf{Cross-validation}. The reader implements $k$-fold
cross-validation with $k = 5$ or $k = 10$, computing the held-out
mean squared error for the regression case and the held-out
log-loss or misclassification rate for the logistic case. The
report includes the mean and the standard error of the
cross-validated metric across folds. Every operation that adapts
to the data, including standardisation of predictors and
hyperparameter tuning, sits inside the cross-validation loop.

\textbf{Variable-selection or regularisation choice}. The reader
makes one explicit modelling choice (which predictors to include,
which transformation to apply, whether to use ridge or lasso,
which value of $\lambda$ to pick) and reports what changes if the
choice is reversed. The report names the cross-validated metric
under both choices.

\textbf{Reporting}. The reader produces:

\begin{itemize}[leftmargin=*]
  \item A table of fitted coefficients with standard errors (OLS
    and logistic) or coefficient values (ridge and lasso).
  \item A residual-diagnostic plot (residuals against fitted
    values for regression; calibration plot or ROC curve for
    logistic).
  \item The cross-validated metric, with the mean and the
    standard error across folds, and a comparison to the
    in-sample metric on the full data.
  \item A statement of which predictors the model relies on most
    and which the cross-validation suggests are not contributing.
  \item A statement of the model's defensible domain (the range of
    predictor values over which the cross-validation is informative)
    and what the model cannot defend.
\end{itemize}

\textbf{Reflection (1500 words)}. The reader writes a 1500-word
narrative addressing:

\begin{itemize}[leftmargin=*]
  \item Why cross-validated error differs from in-sample error on
    this dataset, with a quantitative comparison of the two.
  \item Which residual or calibration anomaly the diagnostic plot
    revealed, and what model misspecification it suggests.
  \item The variable-selection or regularisation choice the reader
    made, and what it would take to reverse the choice. The
    reflection takes the trade-off seriously as a modelling
    decision, not a tuning detail.
  \item One claim the model could be used to defend in a working
    engineering report and one claim it could not, in light of
    the cross-validation, the residuals, and the dataset's
    domain.
  \item What additional data, what additional predictor, or what
    additional pre-registration discipline would tighten the
    report.
\end{itemize}

\textbf{Deliverable}. The dataset (or its citation), the code, the
fitted-coefficient table, the diagnostic plot, the cross-validated
metric with its standard error, and the 1500-word reflection. The
reader deposits the dataset and code at \repopath{volumes/02-form/ch14-statistical-inference/code/}
or links to a stable online location. The general editor's
reference solution on a public dataset is available online; the
reader compares only after completing their own.

\textbf{Time}. Approximately one hour to choose and clean the
dataset, two hours to fit the model and run cross-validation, one
hour to produce the table and diagnostic plot, and three to four
hours to write the reflection. Total approximately seven to eight
hours.
\end{project}

\section{Exercises}

The exercises test calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment. The editor
should attach the solution and rubric file before release.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
For a sample of size $n = 25$ from a Gaussian population with
sample mean $\bar{x} = 12.4$ and sample standard deviation $s =
2.1$, compute the $95\,\%$ confidence interval for $\mu$ and the
$95\,\%$ prediction interval for the next observation. Use
$t_{0.975, 24} \approx 2.064$.

\paragraph{Solution sketch.} The standard error of the mean is
$s/\sqrt{n} = 2.1/5 = 0.42$. The confidence half-width is
$2.064 \times 0.42 \approx 0.87$, giving the interval
$12.4 \pm 0.87 = [11.53, 13.27]$. The prediction-interval inner
factor is $s\sqrt{1 + 1/n} = 2.1\sqrt{1.04} \approx 2.14$, so the
half-width is $2.064 \times 2.14 \approx 4.42$, giving
$12.4 \pm 4.42 = [7.98, 16.82]$. The prediction interval is
roughly $\sqrt{n}$ times wider, which is the working mnemonic.
\end{exercise}

\begin{exercise}
A coin is flipped $n = 40$ times and lands heads $s = 27$ times.
Compute the maximum-likelihood estimate of $p$, its observed-Fisher
standard error, and a $95\,\%$ Wald confidence interval. State the
$z$-quantile used.

\paragraph{Solution sketch.} $\hat{p} = s/n = 27/40 = 0.675$. The
observed Fisher information at $\hat{p}$ is $n/(\hat{p}(1-\hat{p}))
= 40/(0.675 \times 0.325) \approx 182.4$, so the standard error is
$1/\sqrt{182.4} \approx 0.0740$. With $z_{0.975} = 1.96$, the Wald
interval is $0.675 \pm 1.96 \times 0.0740 = [0.530, 0.820]$. The
Wald interval is mildly anti-conservative near the boundaries of
$[0, 1]$; the Wilson or Agresti-Coull interval would tighten near
the endpoints.
\end{exercise}

\begin{exercise}
For an exponential sample with observations $\set*{1.2, 0.5, 2.1,
0.8, 1.7, 0.3, 1.0, 1.5, 0.9, 2.4}$, compute the MLE
$\hat{\lambda} = 1/\bar{x}$. Report $\bar{x}$, $\hat{\lambda}$, and
the observed log-likelihood $\ell(\hat{\lambda})$.

\paragraph{Solution sketch.} Sum is $12.4$, so $\bar{x} = 1.24$
and $\hat{\lambda} = 1/1.24 \approx 0.806$. The log-likelihood at
the MLE is $\ell(\hat{\lambda}) = n\log\hat{\lambda} -
\hat{\lambda}\sum x_{i} = 10\log(0.806) - 0.806 \times 12.4 =
10(-0.2158) - 10 = -2.158 - 10 = -12.16$. Equivalently
$n(\log\hat{\lambda} - 1)$, which is the working closed form for
the exponential MLE's maximised log-likelihood.
\end{exercise}

\begin{exercise}
A Beta$(2, 2)$ prior is placed on the success probability $p$ of a
Bernoulli model. After observing $s = 7$ successes in $n = 10$
trials, compute the posterior parameters and the posterior mean.
Compare the posterior mean to the MLE.

\paragraph{Solution sketch.} The conjugate update gives
$\mathrm{Beta}(\alpha + s, \beta + n - s) = \mathrm{Beta}(9, 5)$.
The posterior mean is $9/(9+5) = 9/14 \approx 0.643$. The MLE is
$\hat{p} = 7/10 = 0.700$. The posterior mean is shrunk toward the
prior mean $0.5$; the shrinkage is the prior's effective sample
size $\alpha + \beta = 4$ acting on a data sample of $n = 10$,
giving a $4/14 \approx 29\,\%$ weight on the prior.
\end{exercise}

\begin{exercise}
For the four data points $(0,1), (1,3), (2,4), (3,7)$, set up the
design matrix $\mat{X}$ for a linear regression with intercept and
solve the normal equations to obtain $\hat{\beta}_{0}$ and
$\hat{\beta}_{1}$. Verify against the closed-form Volume I Chapter
8 expressions.

\paragraph{Solution sketch.} $\mat{X}$ has columns
$(1,1,1,1)^{\transpose}$ and $(0,1,2,3)^{\transpose}$;
$\vect{y} = (1, 3, 4, 7)^{\transpose}$.
$\mat{X}^{\transpose}\mat{X} = \bigl(\begin{smallmatrix} 4 & 6 \\ 6 & 14 \end{smallmatrix}\bigr)$
and $\mat{X}^{\transpose}\vect{y} = (15, 33)^{\transpose}$. The
determinant is $4 \cdot 14 - 36 = 20$, so
$(\mat{X}^{\transpose}\mat{X})^{-1} = \frac{1}{20}\bigl(\begin{smallmatrix} 14 & -6 \\ -6 & 4 \end{smallmatrix}\bigr)$.
Multiplying gives $\hat{\beta}_{0} = (14 \cdot 15 - 6 \cdot
33)/20 = (210 - 198)/20 = 0.6$ and $\hat{\beta}_{1} = (-6 \cdot
15 + 4 \cdot 33)/20 = (-90 + 132)/20 = 2.1$. The Volume I formulae
$\hat{\beta}_{1} = S_{xy}/S_{xx}$ with $\bar{x} = 1.5$,
$\bar{y} = 3.75$, $S_{xy} = 10.5$, $S_{xx} = 5$ give the same
$\hat{\beta}_{1} = 2.1$ and $\hat{\beta}_{0} = 3.75 - 2.1 \times
1.5 = 0.6$.
\end{exercise}

\begin{exercise}
For the Gaussian-Gaussian model with prior $\mu \sim \mathcal{N}(0,
1)$ and known $\sigma^{2} = 4$, compute the posterior mean and
posterior variance after observing the sample $\set*{1.1, 0.7,
2.3}$. State the prior and data precisions used.

\paragraph{Solution sketch.} Prior precision $\tau_{0}^{-2} = 1$;
data precision $n\sigma^{-2} = 3/4 = 0.75$. Posterior precision
$\tau_{n}^{-2} = 1 + 0.75 = 1.75$, so $\tau_{n}^{2} = 4/7
\approx 0.571$. Sample mean $\bar{x} = (1.1 + 0.7 + 2.3)/3 =
1.367$. Posterior mean $\mu_{n} = (4/7)(1 \cdot 0 + 0.75 \cdot
1.367) = (4/7)(1.025) \approx 0.586$. The posterior is
$\mathcal{N}(0.586, 0.571)$. The data have pulled the prior toward
the sample mean by a fraction $0.75/1.75 \approx 43\,\%$.
\end{exercise}

\begin{exercise}
Compute the Bonferroni-corrected significance level for a family of
$m = 12$ simultaneous hypothesis tests at family-wise level $\alpha
= 0.05$. State which individual $p$-values would be rejected from
the list $\set*{0.001, 0.003, 0.012, 0.025, 0.040}$.

\paragraph{Solution sketch.} The Bonferroni threshold is
$\alpha/m = 0.05/12 \approx 0.00417$. Of the five listed
$p$-values, only $0.001$ and $0.003$ are below the threshold; the
other three are not rejected. Note the gap: $0.012$, which would
have been significant at uncorrected $\alpha = 0.05$, fails the
corrected threshold by a factor of nearly three. The
Benjamini-Hochberg procedure applied to the same list with
$m = 12$ would reject $\{p_{(k)} \leq k\alpha/m\}$ for the
largest such $k$; here $p_{(1)} = 0.001 \leq 0.00417$,
$p_{(2)} = 0.003 \leq 0.00833$, $p_{(3)} = 0.012 \leq 0.01250$,
$p_{(4)} = 0.025 > 0.01667$, so BH rejects the first three.
\end{exercise}

\begin{exercise}
For the ridge regression estimator with $\mat{X}^{\transpose}\mat{X}
= \mat{I}_{2}$, $\mat{X}^{\transpose}\vect{Y} = (3, 2)^{\transpose}$,
compute $\hat{\vect{\beta}}_{\mathrm{ridge}}(\lambda)$ for $\lambda
\in \set*{0, 1, 4, 9}$. Plot the coefficient values as a function
of $\lambda$ and identify the limiting behaviour.

\paragraph{Solution sketch.} With orthogonal design,
$\hat{\vect{\beta}}_{\mathrm{ridge}}(\lambda) = (1+\lambda)^{-1}
\mat{X}^{\transpose}\vect{Y}$. Evaluating at the four
$\lambda$-values gives $\hat{\vect{\beta}}(0) = (3, 2)$,
$\hat{\vect{\beta}}(1) = (1.5, 1.0)$, $\hat{\vect{\beta}}(4) =
(0.6, 0.4)$, $\hat{\vect{\beta}}(9) = (0.3, 0.2)$. Both
coordinates shrink toward zero by the common factor
$1/(1+\lambda)$; the relative direction $(3:2)$ is preserved.
The limiting behaviour as $\lambda \to \infty$ is
$\hat{\vect{\beta}} \to \vect{0}$.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Show that the sample variance with the Bessel correction,
$S_{n}^{2} = \frac{1}{n-1}\sum (X_{i} - \bar{X}_{n})^{2}$, is
unbiased for the population variance $\sigma^{2}$. State the
expectation identity used at each step.

\paragraph{Solution sketch.} Expand
$\sum_{i}(X_{i} - \bar{X})^{2} = \sum_{i} X_{i}^{2} - n\bar{X}^{2}$.
Taking expectations and using $\E[X_{i}^{2}] = \mu^{2} + \sigma^{2}$
and $\E[\bar{X}^{2}] = \mu^{2} + \sigma^{2}/n$ gives
$\E[\sum (X_{i} - \bar{X})^{2}] = n(\mu^{2} + \sigma^{2}) - n(\mu^{2}
+ \sigma^{2}/n) = (n-1)\sigma^{2}$. Dividing by $n-1$ yields
$\E[S_{n}^{2}] = \sigma^{2}$. The Bessel correction recovers the
single degree of freedom consumed by estimating $\mu$ with
$\bar{X}$.
\end{exercise}

\begin{exercise}
Derive the bias-variance decomposition $\operatorname{MSE}(\hat{\theta})
= \Var(\hat{\theta}) + \operatorname{bias}(\hat{\theta})^{2}$ from
the definition $\operatorname{MSE}(\hat{\theta}) = \E[(\hat{\theta}
- \theta)^{2}]$, using the addition and subtraction of
$\E[\hat{\theta}]$.

\paragraph{Solution sketch.} Write
$\hat{\theta} - \theta = (\hat{\theta} - \E\hat{\theta}) +
(\E\hat{\theta} - \theta)$. Squaring and taking expectations,
the cross term $2\E[(\hat{\theta} - \E\hat{\theta})(\E\hat{\theta} -
\theta)]$ vanishes because $\E\hat{\theta} - \theta$ is constant
and $\E[\hat{\theta} - \E\hat{\theta}] = 0$. The remaining two
terms are $\E[(\hat{\theta} - \E\hat{\theta})^{2}] =
\Var(\hat{\theta})$ and $(\E\hat{\theta} - \theta)^{2} =
\operatorname{bias}(\hat{\theta})^{2}$.
\end{exercise}

\begin{exercise}
Derive the MLE for a Poisson sample $X_{1}, \ldots, X_{n}$. Compute
its observed Fisher information at $\hat{\lambda}$ and the
asymptotic variance. State the regularity condition you used to
move the derivative inside the expectation.

\paragraph{Solution sketch.} The log-likelihood is
$\ell(\lambda) = (\sum x_{i})\log\lambda - n\lambda + \text{const}$.
Setting $\dd\ell/\dd\lambda = (\sum x_{i})/\lambda - n = 0$ gives
$\hat{\lambda} = \bar{x}$. The second derivative is
$-(\sum x_{i})/\lambda^{2}$; its negation, evaluated at
$\hat{\lambda}$, is the observed information $n/\bar{x}$.
$I_{n}(\lambda) = n/\lambda$ in expectation, so the asymptotic
variance per observation is $\lambda$; the standard error of
$\hat{\lambda}$ is $\sqrt{\bar{x}/n}$. The regularity condition
used is that the support $\set{0, 1, 2, \ldots}$ does not depend
on $\lambda$, allowing the dominated-convergence interchange of
derivative and expectation.
\end{exercise}

\begin{exercise}
Show that, for the Beta-Bernoulli conjugate model with prior
$\mathrm{Beta}(\alpha, \beta)$ and data with $s$ successes in $n$
trials, the posterior is $\mathrm{Beta}(\alpha + s, \beta + n - s)$.
Identify the prior parameters as ``prior successes'' and ``prior
failures'' and discuss the working intuition.

\paragraph{Solution sketch.} The posterior is proportional to
$L(p) \pi(p) \propto p^{s}(1-p)^{n-s} \cdot p^{\alpha-1}
(1-p)^{\beta-1} = p^{\alpha+s-1}(1-p)^{\beta+n-s-1}$, the kernel
of $\mathrm{Beta}(\alpha + s, \beta + n - s)$. The prior
parameters $\alpha$ and $\beta$ act as prior pseudo-counts: a
$\mathrm{Beta}(\alpha, \beta)$ prior has the inferential weight of
$\alpha + \beta$ prior observations, of which $\alpha$ are prior
successes. The posterior mean $(\alpha + s)/(\alpha + \beta + n)$
is a precision-weighted average of the prior mean $\alpha/
(\alpha+\beta)$ and the MLE $s/n$.
\end{exercise}

\begin{exercise}
Show that, in the Gaussian linear regression model with full-rank
design matrix $\mat{X}$, the OLS estimator satisfies
$\E[\hat{\vect{\beta}}] = \vect{\beta}$ and $\Var(\hat{\vect{\beta}})
= \sigma^{2}(\mat{X}^{\transpose}\mat{X})^{-1}$. State, but do not
prove, the Gauss-Markov theorem.

\paragraph{Solution sketch.} Substitute $\vect{Y} = \mat{X}
\vect{\beta} + \vect{\varepsilon}$ into the closed form:
$\hat{\vect{\beta}} = (\mat{X}^{\transpose}\mat{X})^{-1}
\mat{X}^{\transpose}(\mat{X}\vect{\beta} + \vect{\varepsilon}) =
\vect{\beta} + (\mat{X}^{\transpose}\mat{X})^{-1}
\mat{X}^{\transpose}\vect{\varepsilon}$. Taking expectations and
using $\E[\vect{\varepsilon}] = \vect{0}$ gives unbiasedness.
$\Var(\hat{\vect{\beta}}) = (\mat{X}^{\transpose}\mat{X})^{-1}
\mat{X}^{\transpose}\Var(\vect{\varepsilon})\mat{X}
(\mat{X}^{\transpose}\mat{X})^{-1} = \sigma^{2}(\mat{X}^{\transpose}
\mat{X})^{-1}$, using
$\Var(\vect{\varepsilon}) = \sigma^{2}\mat{I}$ and the
symmetry of $\mat{X}^{\transpose}\mat{X}$. The Gauss-Markov
theorem: among all linear unbiased estimators of $\vect{\beta}$,
the OLS estimator has the smallest variance in the matrix
positive-semidefinite ordering.
\end{exercise}

\begin{exercise}
Derive the closed form $\hat{\vect{\beta}}_{\mathrm{ridge}}(\lambda)
= (\mat{X}^{\transpose}\mat{X} + \lambda \mat{I})^{-1}
\mat{X}^{\transpose}\vect{Y}$ from the optimisation problem
$\min_{\vect{\beta}} \norm*{\vect{Y} - \mat{X}\vect{\beta}}_{2}^{2}
+ \lambda \norm*{\vect{\beta}}_{2}^{2}$ by setting the gradient
with respect to $\vect{\beta}$ to zero.

\paragraph{Solution sketch.} Expand the objective as
$\vect{Y}^{\transpose}\vect{Y} - 2\vect{\beta}^{\transpose}
\mat{X}^{\transpose}\vect{Y} + \vect{\beta}^{\transpose}
\mat{X}^{\transpose}\mat{X}\vect{\beta} + \lambda
\vect{\beta}^{\transpose}\vect{\beta}$. Differentiating with
respect to $\vect{\beta}$ and setting to zero:
$-2\mat{X}^{\transpose}\vect{Y} + 2(\mat{X}^{\transpose}\mat{X}
+ \lambda \mat{I})\vect{\beta} = \vect{0}$. Rearranging gives
the claimed closed form. The matrix $\mat{X}^{\transpose}\mat{X}
+ \lambda\mat{I}$ is positive definite for $\lambda > 0$ even
when $\mat{X}^{\transpose}\mat{X}$ is singular, which is the
operational reason ridge always has a unique solution.
\end{exercise}

\begin{exercise}
Show that, for a generalised likelihood-ratio test under regularity
conditions, the test statistic $-2\ln\Lambda(X)$ has an asymptotic
$\chi^{2}_{r}$ distribution under the null, where $r$ is the
difference in dimension between the alternative and the null
parameter sets. Sketch the argument; do not work the regularity
conditions.

\paragraph{Solution sketch.} Let $\hat{\theta}$ be the
unrestricted MLE on $\Theta$ and $\hat{\theta}_{0}$ the
restricted MLE on $\Theta_{0}$. Taylor-expand the log-likelihood
$\ell$ around $\hat{\theta}$ to second order:
$\ell(\hat{\theta}_{0}) \approx \ell(\hat{\theta}) - \frac{1}{2}
(\hat{\theta} - \hat{\theta}_{0})^{\transpose} I_{n}(\hat{\theta})
(\hat{\theta} - \hat{\theta}_{0})$, since the first-order term
vanishes at the MLE. Multiplying through by $-2$,
$-2\ln\Lambda = -2[\ell(\hat{\theta}_{0}) - \ell(\hat{\theta})]
\approx (\hat{\theta} - \hat{\theta}_{0})^{\transpose}
I_{n}(\hat{\theta}) (\hat{\theta} - \hat{\theta}_{0})$. Under
$H_{0}$, $\sqrt{n}(\hat{\theta} - \hat{\theta}_{0})$ is
asymptotically Gaussian with covariance $I_{1}(\theta)^{-1}$ on
the $r$-dimensional complement of $\Theta_{0}$ in $\Theta$, so
the quadratic form has an asymptotic $\chi^{2}_{r}$ distribution.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the sample size required to estimate a population mean to
within $\pm 1\,\%$ relative precision at $95\,\%$ confidence, given
a working coefficient of variation $\sigma/\mu = 0.1$. State the
working approximation used.

\paragraph{Solution sketch.} The half-width of the $95\,\%$
confidence interval is $1.96 \sigma/\sqrt{n}$. Setting this equal
to $0.01 \mu$ and using $\sigma = 0.1\mu$ gives $1.96 \times
0.1\mu /\sqrt{n} = 0.01\mu$, so $\sqrt{n} = 19.6$, $n \approx
384$. The working rule: at $95\,\%$ confidence, the sample size
scales as $(z_{1-\alpha/2})^{2}(\sigma/\mu)^{2}/r^{2}$ where $r$
is the relative precision; here $(1.96)^{2}(0.1)^{2}/(0.01)^{2}
\approx 384$.
\end{exercise}

\begin{exercise}
Estimate the family-wise false-positive rate when an analyst runs
$m = 20$ independent tests at $\alpha = 0.05$ each and reports the
smallest $p$-value as a single test result. State the approximate
formula and the result to two significant figures.

\paragraph{Solution sketch.} The probability that at least one of
$m = 20$ independent tests crosses the $\alpha = 0.05$ threshold
under the global null is $1 - (1 - 0.05)^{20} = 1 - (0.95)^{20}$.
With $(0.95)^{20} \approx 0.358$, the family-wise rate is
approximately $0.64$, or $64\,\%$. A nominally $5\,\%$ test
reported as a single result from $20$ trials is in fact about a
$64\,\%$ test.
\end{exercise}

\begin{exercise}
Estimate the half-width of a $95\,\%$ Bayesian credible interval
for a Bernoulli success probability under a Beta$(2, 2)$ prior
after $n = 50$ observations, using the working approximation that
the Beta posterior is approximately Gaussian for moderate
parameters. Compare to the frequentist Wald interval.

\paragraph{Solution sketch.} Suppose $s = 30$ successes. The
posterior is $\mathrm{Beta}(32, 22)$ with posterior mean $32/54
\approx 0.593$ and posterior variance $\alpha'\beta'/((\alpha' +
\beta')^{2}(\alpha' + \beta' + 1)) = 32 \cdot 22 / (54^{2} \cdot
55) \approx 0.00439$, so posterior standard deviation $\approx
0.0663$. The $95\,\%$ credible half-width is approximately
$1.96 \times 0.0663 \approx 0.130$. The Wald frequentist interval
uses $\hat{p}(1-\hat{p})/n$ with $\hat{p} = 0.6$: standard error
$\sqrt{0.24/50} \approx 0.0693$, half-width $\approx 0.136$.
The two are close because the prior is weak ($\alpha + \beta = 4$
compared to $n = 50$).
\end{exercise}

\begin{exercise}
Estimate the cross-validated standard error of a $5$-fold
cross-validated mean squared error for a linear regression with
$n = 100$ observations, given an in-sample residual standard
deviation of $\sigma_{\hat{\varepsilon}} = 0.5$. State the working
approximation.

\paragraph{Solution sketch.} The per-fold mean squared error
estimate has variance approximately $2\sigma^{4}/n_{\mathrm{fold}}$
under Gaussian residuals; with $n_{\mathrm{fold}} = 20$ and
$\sigma^{2} = 0.25$, the per-fold variance is about $2(0.25)^{2}/
20 = 0.00625$, so per-fold standard deviation $\approx 0.079$.
Averaging across $k = 5$ folds and treating them as approximately
independent gives a standard error of the cross-validated metric
of $0.079/\sqrt{5} \approx 0.035$. The folds are not fully
independent (the training sets overlap), so the actual standard
error is somewhat larger; the working approximation is treated
as a lower bound, and the engineer reports the empirical
across-fold standard deviation alongside.
\end{exercise}

\begin{exercise}
Estimate the value of the regularisation parameter $\lambda$ that
makes a ridge regression's effective degrees of freedom (the trace
of the hat matrix) equal to half the rank $p$ of the design
matrix, when $\mat{X}^{\transpose}\mat{X} = \mat{I}_{p}$. State the
formula used.

\paragraph{Solution sketch.} For ridge with orthogonal design,
the hat matrix is $\mat{H} = \mat{X}(\mat{I} + \lambda\mat{I})^{-1}
\mat{X}^{\transpose}$, and the effective degrees of freedom are
$\tr(\mat{H}) = p/(1 + \lambda)$. Setting $p/(1+\lambda) = p/2$
gives $\lambda = 1$. In words, $\lambda$ equal to the
eigenvalue of $\mat{X}^{\transpose}\mat{X}$ halves the effective
dimensionality of the model. For a general design, replace
$\lambda$ by the working ratio of $\lambda$ to the typical
eigenvalue of $\mat{X}^{\transpose}\mat{X}$.
\end{exercise}

\begin{exercise}
Estimate the generalisation error inflation factor for a model
with $p = 20$ parameters fit to $n = 100$ observations, given an
in-sample residual mean squared error of $\sigma^{2}_{\hat{\varepsilon}}$.
Use Mallows' $C_{p}$ approximation as the working rule.

\paragraph{Solution sketch.} The expected generalisation error
under Mallows' approximation is $\sigma^{2}(1 + 2p/n)$, where
$\sigma^{2}$ is the true noise variance, estimated by
$\sigma^{2}_{\hat{\varepsilon}}\,n/(n-p)$ to correct the in-sample
estimate. With $p = 20$, $n = 100$, the in-sample inflation
factor is $1 + 2(20)/100 = 1.40$. The $40\,\%$ inflation is the
canonical reason an in-sample mean squared error from a
$20$-parameter model on $100$ observations is not a defensible
estimate of generalisation error; cross-validation or hold-out is
required.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
In a programming environment of the reader's choice, simulate
$10\,000$ samples of size $n = 30$ from a Gaussian population with
$\mu = 10$, $\sigma = 2$. For each sample, compute the $95\,\%$
$t$-confidence interval for $\mu$. Verify that the empirical
coverage is approximately $0.95$ and report the sample standard
error of the empirical coverage rate.

\paragraph{Solution sketch.} The expected coverage is exactly
$0.95$ under the Gaussian model. The standard error of the
empirical coverage rate, treated as a binomial proportion with
$N = 10\,000$, is $\sqrt{0.95 \times 0.05 / 10000} \approx
0.0022$, so the empirical coverage should land within roughly
$0.945$ to $0.955$ on a typical run. A seed-fixed implementation
with $t_{0.975, 29} \approx 2.045$ produces a value in that band;
a value outside the band on a single run signals an
implementation bug, not a Monte Carlo failure.
\end{exercise}

\begin{exercise}
Simulate the p-hacking inflation of section 14.7. Generate $1000$
trials in which a true effect of zero exists, run $m = 5$
$t$-tests on different post hoc subgroups (random partitions),
and report the smallest $p$-value as a single result. Verify that
the empirical false-positive rate at nominal $\alpha = 0.05$
exceeds $0.05$ and approaches the analytical $1 - (1-\alpha)^{m}$.

\paragraph{Solution sketch.} The analytical false-positive rate
under independent tests is $1 - (1 - 0.05)^{5} \approx 0.226$.
Under random subgroup partitions of the same data, the tests are
positively dependent, so the empirical inflation will land
somewhere between $0.05$ (perfectly correlated tests) and $0.226$
(independent tests). A typical implementation produces an
empirical rate near $0.16$ to $0.20$, which is enough to
demonstrate the inflation and motivate the multiple-comparisons
correction. The standard error of the empirical rate at
$N = 1000$ is approximately $\sqrt{0.2 \times 0.8 / 1000}
\approx 0.013$.
\end{exercise}

\begin{exercise}
Implement $5$-fold cross-validation of a polynomial regression of
degree $p \in \set*{1, 2, \ldots, 8}$ on $n = 60$ points generated
from $y = \sin(2\pi x) + \varepsilon$ with $\varepsilon \sim
\mathcal{N}(0, 0.2^{2})$ and $x \in [0, 1]$. Plot the training
mean squared error and the cross-validated mean squared error
against $p$. Identify the degree at which the two curves diverge.

\paragraph{Solution sketch.} Training mean squared error
decreases monotonically with $p$ (the higher-order model contains
the lower-order one). The cross-validated curve typically
reaches a minimum near $p = 3$ to $p = 5$ (a cubic or quartic
captures a half-period of the sine) and rises sharply for $p
\geq 6$ as the model begins to fit the noise. The divergence
point is the operational definition of overfitting under this
loss; the engineer reports the cross-validation curve, not the
training curve, when defending a model-complexity choice.
\end{exercise}

\begin{exercise}
Simulate the bias-variance trade-off of ridge regression. Generate
$200$ datasets of size $n = 50$ from a linear model with $p = 10$
predictors and $\sigma = 1$. For each dataset, fit ridge with
$\lambda \in \set*{0, 0.1, 1, 10, 100}$. Compute the empirical
bias squared and variance of $\hat{\vect{\beta}}_{1}$ across the
$200$ datasets at each $\lambda$. Verify that the MSE has a
minimum at intermediate $\lambda$.

\paragraph{Solution sketch.} Variance of
$\hat{\vect{\beta}}_{1}$ across the $200$ runs decreases
monotonically with $\lambda$. Squared bias (squared difference
between the run-averaged $\hat{\vect{\beta}}_{1}$ and the true
$\beta_{1}$) increases monotonically. The sum, the empirical
mean squared error, displays the U-shape sketched in
figure~\ref{fig:vol02:ch14:bias-variance}: it is high at $\lambda
= 0$ (high variance from the near-singular design), reaches a
minimum near $\lambda = 1$ to $\lambda = 10$ depending on the
design's conditioning, and rises again at $\lambda = 100$ as
the bias dominates. The minimum is broad; the precise value of
$\lambda$ that achieves it depends on the seed.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a hypothesis test for a manufacturing process whose target
mean output is $\mu_{0} = 100\,\si{\gram}$. State the null and
alternative, the test statistic, the significance level chosen
in light of the cost of a false alarm and a missed shift, the
sample size required to detect a shift of $\delta = 2\,\si{\gram}$
with power $0.90$ given $\sigma = 5\,\si{\gram}$, and the decision
rule. Justify each choice.

\paragraph{Discussion.} A defensible design uses a two-sided
test $H_{0}: \mu = 100$ versus $H_{1}: \mu \neq 100$ with
statistic $Z = (\bar{X} - 100)/(\sigma/\sqrt{n})$. The
significance level depends on cost: a setting where a false alarm
triggers an inexpensive recalibration tolerates $\alpha = 0.05$;
a setting where it triggers a costly line stop justifies
$\alpha = 0.01$. Sample size at $\alpha = 0.05$, $1-\beta =
0.90$ uses $n \approx ((z_{0.975} + z_{0.90})\sigma/\delta)^{2}
= ((1.96 + 1.28) \times 5 / 2)^{2} \approx 66$. Decision rule:
reject $H_{0}$ if $\abs*{Z} > 1.96$, with the test conducted on
$n = 66$ consecutive units. The design's justification names
both errors and reports the power against the
engineering-relevant shift, not just the significance.
\end{exercise}

\begin{exercise}
Design a Bayesian update procedure for a quality-control inspector
who tests batches of components for a defect probability $p$. State
the prior, the likelihood, the posterior update rule per batch,
and the reporting threshold (the posterior probability of a defect
rate above some critical level $p^{\star}$ that triggers a process
halt). Justify the prior in terms of working knowledge of the
process.

\paragraph{Discussion.} A working design uses prior $p \sim
\mathrm{Beta}(\alpha_{0}, \beta_{0})$, with $\alpha_{0}, \beta_{0}$
chosen so the prior mean matches the historical defect rate (for
instance $\alpha_{0} = 2$, $\beta_{0} = 198$ for a $1\,\%$ prior
on $n = 200$ historical units). After a batch of $n_{b}$ tests
with $s_{b}$ defects, the posterior is $\mathrm{Beta}(\alpha_{0}
+ s_{\text{cum}}, \beta_{0} + n_{\text{cum}} - s_{\text{cum}})$
with cumulative counts across all batches. The reporting
threshold is the posterior probability $\Pr[p > p^{\star} \mid
\text{data}]$ exceeding a chosen level (e.g.\ $0.10$); when
exceeded, the inspector halts the process. The prior must be
defensible from prior production runs; a flat $\mathrm{Beta}(1,1)$
prior is rarely defensible in a mature process and would treat
the first defect as far more diagnostic than working knowledge
warrants.
\end{exercise}

\begin{exercise}
Design a $k$-fold cross-validation procedure for a logistic
regression of failure on three sensor readings. State $k$, the
loss function, the pre-processing operations and which sit inside
the loop, the hyperparameter search grid, and the reporting
metric. State the rule by which the cross-validation result is
trusted as an estimate of generalisation.

\paragraph{Discussion.} A defensible design uses stratified
$k = 5$ folds to preserve the failure-rate balance per fold.
Loss is log-loss (proper scoring rule) with a calibration plot
as a diagnostic. Pre-processing inside the loop: per-fold
standardisation of the three sensor readings, per-fold imputation
of any missing values. Outside the loop: time-invariant
operations such as variable renaming and physical-unit
conversion. The hyperparameter search grid is the $\ell_{2}$
regularisation strength $C \in \set{0.01, 0.1, 1, 10, 100}$,
chosen by inner cross-validation if the engineer wants a
fully nested procedure. Reporting metric: mean and across-fold
standard error of held-out log-loss, plus held-out classification
accuracy at a stated decision threshold. The procedure is
trusted as an estimate of generalisation when no operation
adapts to the data outside the loop and when the failure rate
is preserved across folds; if the failure rate is below
$5\,\%$, stratification or a different evaluation metric (such
as AUC) is required.
\end{exercise}

\begin{exercise}
Design a multiple-comparisons-controlled analysis plan for a
study that runs $m = 50$ simultaneous hypothesis tests on a
genomics-style dataset. State the family-wise control method
(Bonferroni) versus the false-discovery-rate control method
(Benjamini-Hochberg), the significance level chosen, the criterion
by which the analyst decides between the two methods, and the
report format.

\paragraph{Discussion.} The choice between Bonferroni and
Benjamini-Hochberg follows the cost structure. Bonferroni
controls the family-wise error rate at the cost of low power
when $m$ is large; it is appropriate when a single false rejection
is itself costly (a confirmatory finding that triggers expensive
follow-up). Benjamini-Hochberg controls the expected proportion
of false rejections among rejections; it is appropriate when
the cost of a missed true rejection outweighs the cost of one
or two false ones (exploratory genomics). At $m = 50$ and
$\alpha = 0.05$, Bonferroni rejects at per-test threshold
$0.001$; BH rejects up to the largest $k$ with $p_{(k)} \leq k
\times 0.001$. The report names the chosen method, the
per-comparison threshold, the number of tests run, the number
rejected, and the false-discovery-rate or family-wise-error
guarantee. The unreported $m$ is the disclosure failure that
inflates the literature
\cite{method:simmons-nelson-simonsohn-2011}.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A colleague reports a regression with in-sample $R^{2} = 0.99$ and
no cross-validated error. Identify the three things that should
make the engineer suspicious of the report and propose follow-up
questions.

\paragraph{Solution sketch.} Three suspicions are warranted.
First, the in-sample $R^{2}$ is large because the model has many
parameters relative to $n$, and the cross-validated $R^{2}$ may
be far smaller. Ask for $n$, $p$, and the cross-validated
metric. Second, the predictors may include leakage (a variable
that is a proxy for or downstream of the response). Ask for the
full predictor list and the temporal or causal relationship
between each predictor and the response. Third, the response
variable may have been transformed in a way that inflates
$R^{2}$ artificially (logging a heavily skewed variable can
move $R^{2}$ from $0.5$ to $0.95$ without improving prediction
of the original quantity). Ask for the working units of the
response and the metric reported on the original scale.
\end{exercise}

\begin{exercise}
A colleague reports a $p$-value of $0.03$ for a single test from a
study that examined ``several outcomes.'' Identify the additional
information the engineer needs to interpret the $p$-value and the
correction that would bring the report into the working discipline.

\paragraph{Solution sketch.} The engineer needs $m$ (the number
of outcomes examined), whether the analysis was pre-registered,
and which outcomes were planned versus selected after the data
were seen. With $m = 5$ untargeted tests at nominal $\alpha =
0.05$, the family-wise rate is $1 - 0.95^{5} \approx 0.23$, and
the reported $p = 0.03$ is consistent with a global null. The
correction depends on the framing: Bonferroni gives a corrected
$p$-value of $\min(5 \times 0.03, 1) = 0.15$, no longer
significant at $\alpha = 0.05$; Benjamini-Hochberg at $\alpha =
0.05$ requires $p_{(k)} \leq k \times 0.01$ for the ranked
list. If the study was not pre-registered and the chosen
outcome was selected for promise, the correct interpretation is
``working hypothesis to be replicated,'' not ``established
effect.''
\end{exercise}

\begin{exercise}
A logistic regression's residual deviance is far smaller than the
degrees of freedom would predict under the model. Diagnose the
two most likely causes (over-fit due to too many predictors;
contamination of the training set with test data) and propose the
diagnostic that distinguishes the two.

\paragraph{Solution sketch.} Under a correctly specified
logistic model, $\E[\text{residual deviance}] \approx
\text{residual degrees of freedom}$. A deviance much smaller than
the degrees of freedom indicates the model is fitting too
closely. Over-fit by too many predictors shows up as a deviance
that recovers to the expected level under $k$-fold cross-validation
(the held-out deviance is much larger than the in-sample). Data
leakage shows up as a deviance that remains small under
cross-validation; the model has access to the response within the
training fold through a leaked variable. The distinguishing
diagnostic: refit on the cross-validation training folds and
compare in-sample versus held-out deviance per fold. A large gap
points to over-fit; a small gap with small overall deviance
points to leakage.
\end{exercise}

\begin{exercise}
A ridge regression chosen by cross-validation produces
$\hat{\vect{\beta}}_{\mathrm{ridge}}$ that disagrees with the OLS
estimator $\hat{\vect{\beta}}_{\mathrm{OLS}}$ in sign on one
coefficient. Diagnose what this sign reversal indicates about the
design matrix's conditioning and the variance of the OLS
coefficient. Propose the diagnostic that confirms the diagnosis.

\paragraph{Solution sketch.} A sign reversal under ridge
indicates that the OLS coefficient was estimated with high
variance, large enough that the sign is not stable across small
perturbations of the data. The underlying cause is collinearity:
two or more predictors carry similar information, the design
matrix has a small eigenvalue, and the corresponding OLS
coefficient absorbs a noisy linear combination that ridge
shrinks toward zero. Confirming diagnostics: compute the
variance inflation factors (VIF) of each predictor; a VIF above
10 confirms collinearity. Examine the singular values of
$\mat{X}$; a condition number above $30$ indicates an
ill-conditioned design. Refit OLS on bootstrapped resamples and
inspect the resampled distribution of the coefficient; a wide
distribution that straddles zero confirms the OLS estimate's
sign is not informative.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Argue, in 200 to 300 words, why the reproducibility rate reported
in the 2015 Reproducibility Project: Psychology should change the
way an engineer reads a single published study with $p = 0.04$.
Cite \cite{method:open-science-collaboration-2015} and connect the
argument to the disclosure rule of section 14.7.

\paragraph{Solution sketch.} A defensible argument: the Open
Science Collaboration found that only $36\,\%$ of originally
significant findings replicated, with replication effects
averaging half the original. A single published $p = 0.04$ is
therefore consistent with one of two regimes: a true effect
that will replicate at smaller magnitude, or a false-positive
arising from p-hacking, the garden of forking paths, or
publication bias. The published $p$-value alone does not
distinguish the two. The disclosure rule of section 14.7
requires the engineer to ask what tests were run, whether
pre-registration was in place, whether the analysis was the
planned one, and whether the effect size carries a confidence
interval that excludes a working null. A $p = 0.04$ that
survives a multiple-comparisons correction, was specified before
the data were seen, comes from a high-powered design, and is
accompanied by an effect-size interval that excludes negligible
effects is a different working object from a $p = 0.04$ that
appears without context. The engineer who sizes a load-bearing
design against the second class of evidence is treating a single
observation as confirmation, which the empirical replication
record warns against. The working compromise is to read a
single $p = 0.04$ as a working hypothesis to be replicated, not
as a result to be acted on, until the disclosures are made and
the effect survives an honest correction.
\end{exercise}

\begin{exercise}
Argue, in 200 to 300 words, why pre-registration alone does not
solve the garden-of-forking-paths problem
\cite{method:gelman-loken-forking-paths}, and identify the
additional discipline that a reader should ask of an analysis to
defend its $p$-value.

\paragraph{Solution sketch.} Pre-registration commits the
analyst to a primary test and a primary outcome before the data
are seen. The forking-paths argument shows that even
pre-registered analyses can be undermined when the
pre-registration is loose or when secondary analyses absorb the
flexibility that the primary analysis was supposed to remove.
A pre-registration that says ``a $t$-test will be conducted''
without specifying covariate adjustment, outlier handling, or
the subgroup of interest still permits substantial post hoc
flexibility. The additional discipline a reader should require:
the pre-registration is specific (test, model, covariates,
outlier rule, subgroup, decision threshold); the analysis report
explicitly enumerates every deviation from the pre-registration
and its justification; every secondary analysis is reported
alongside the primary, with the same level of detail; the
underlying data and analysis code are available for independent
replication. Pre-registration in the absence of these companion
disciplines reduces but does not eliminate the inflation. The
defensible report distinguishes confirmatory results (specified
in advance, all deviations disclosed) from exploratory results
(discovered in the data, reported as hypotheses for replication,
not as established effects).
\end{exercise}

\begin{exercise}
A cross-validated logistic regression on a diagnostic-imaging
dataset shows a held-out accuracy of $0.95$. On a new dataset from
a different hospital, the accuracy drops to $0.71$. Argue, in 200
to 300 words, whether the gap is consistent with overfitting,
with covariate shift, or with leakage of patient-identifier
information into the original training set, and propose the
diagnostic that distinguishes the three.

\paragraph{Solution sketch.} The gap is large enough to demand
investigation, and all three explanations are consistent with
the observation. Overfitting would imply the original
cross-validation was honest but the model has memorised
within-dataset noise that does not generalise. Diagnostic:
re-fit on the original dataset with stronger regularisation and
re-evaluate on the new hospital; if accuracy recovers, the
original was overfit. Covariate shift means the predictor
distribution differs between the two hospitals (different
scanner, different patient population, different acquisition
protocol). Diagnostic: compare the marginal distributions of
each predictor across the two datasets; substantial
distributional differences confirm covariate shift. Patient-
identifier leakage (the model learned to identify individual
patients via incidental features, and the cross-validation folds
contained multiple images per patient) implies the original
cross-validation was contaminated. Diagnostic: split folds by
patient rather than by image and re-run the original
cross-validation; a drop to near $0.71$ confirms leakage. The
three diagnostics together identify the failure mode. The
cross-hospital evaluation is the discipline that surfaces all
three; an engineer who reports only the within-hospital
cross-validation accuracy is reporting a number that does not
defend deployment across institutions.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A research group reports a Bayesian credible interval based on a
strong prior whose source is not stated. Argue, in one paragraph,
what the engineer should ask before treating the interval as
load-bearing for design and what disclosures should accompany
every Bayesian report.

\paragraph{Discussion.} A strong prior with unstated source is
the Bayesian analogue of an undisclosed analysis choice. The
engineer should ask: what is the parametric form of the prior;
what is its effective sample size (the prior's pseudo-count
relative to the data); what evidence underwrites the prior
(earlier studies with citations, physical constraints with
derivations, expert elicitation with the elicitation protocol);
and how does the posterior change under a flat or weakly
informative prior. A Bayesian report that does not disclose the
prior, its effective sample size, the prior-elicitation
provenance, and a sensitivity analysis is not reportable. The
credible interval inherits whatever the prior pulls it toward;
without the disclosures the engineer cannot distinguish a
genuine posterior update from a prior assertion dressed in
data.
\end{exercise}

\begin{exercise}
A vendor reports a logistic regression's classification accuracy
on the data the model was trained on, with no cross-validation.
Argue, in one paragraph, why this is the wrong figure to compare
across vendors and what figure the engineer should request
instead.

\paragraph{Discussion.} In-sample classification accuracy is
inflated by the model's flexibility relative to the dataset.
Across vendors with different model complexities, the more
flexible model wins on in-sample accuracy almost regardless of
its actual generalisation performance; a comparison made on
in-sample accuracy systematically prefers overfit models. The
engineer should request held-out accuracy on a fresh test set
that the vendor never saw during training and hyperparameter
tuning. Better: held-out accuracy on a benchmark dataset that
the engineer holds in escrow and only releases for the
evaluation. Best: held-out accuracy plus a calibration plot and
held-out log-loss; accuracy alone misses the model's confidence
calibration, which matters for downstream cost-sensitive
decisions. The figure to compare is the one the vendor cannot
have optimised against.
\end{exercise}

\begin{exercise}
A statistical consultant proposes choosing the regularisation
parameter $\lambda$ in a ridge regression by minimising training
error. Argue, in one paragraph, why this choice does not control
generalisation error and what the consultant should do instead.
Reference the bias-variance trade-off of section 14.1.

\paragraph{Discussion.} Training error decreases monotonically
as $\lambda \to 0$: with no penalty the OLS fit minimises the
residual sum of squares by construction. Choosing $\lambda$ to
minimise training error therefore always picks $\lambda = 0$,
which is OLS, and forfeits the variance reduction the
bias-variance decomposition promises. The generalisation error,
by contrast, has a minimum at some $\lambda^{\star} > 0$
whenever the design is ill-conditioned (figure
\ref{fig:vol02:ch14:bias-variance}). The consultant should
choose $\lambda$ by minimising the cross-validated error across
a grid (typically logarithmically spaced from $10^{-3}$ to
$10^{3}$ times the typical eigenvalue of $\mat{X}^{\transpose}
\mat{X}$), report the cross-validated curve, and apply the
one-standard-error rule for the final choice. Training error is
the wrong target for any hyperparameter that exists to
trade-off bias against variance, including $\lambda$, polynomial
degree, neural-network width, and regression-tree depth.
\end{exercise}

\begin{exercise}
An analyst presents a study reporting an effect size of $0.6$
standard deviations and a $p$-value of $0.04$ from $n = 24$. The
engineer is asked to use the result as an input to a load-bearing
design decision. Argue, in 200 to 300 words, what additional
evidence the engineer should require before using the result, in
light of the replication-rate evidence in
\cite{method:open-science-collaboration-2015,method:errington-cancer-biology-2021}
and the HARKing literature \cite{method:kerr-harking-1998}.

\paragraph{Discussion.} The numbers describe a low-power
study: at $\alpha = 0.05$ and $d = 0.6$, $n = 24$ gives power
of roughly $0.45$. A barely significant result from a low-power
study, with no replication, is exactly the configuration the
replication-rate evidence flags as most likely to evaporate on
re-analysis: the Open Science Collaboration found replication
effects averaging half the original, the Errington cancer-biology
project found a factor-of-three attenuation, and both studies
emphasised low original power as a predictor of non-replication.
The HARKing literature adds that a single significant result from
a study that examined multiple outcomes may have been selected
after the fact, inflating the apparent effect size. Before
using the result, the engineer should require: the
pre-registration (or its absence and the consequences),
the full list of analyses run, the cross-validated or held-out
effect-size estimate, at least one independent replication, and
the effect-size confidence interval (which at $n = 24$ is wide
enough that the lower bound may include effects too small for
the design to depend on). If no replication exists, the
engineer should size the design against the lower confidence
bound rather than the point estimate, accept that the
load-bearing claim cannot rest on this single study, and
commission or wait for an independent replication. The
disclosure rule of section 14.7 is the working test.
\end{exercise}
