\chapter{Why we measure}
\label{vol01:ch01}

\epigraph{When you can measure what you are speaking about, and express it
in numbers, you know something about it; when you cannot measure it, when
you cannot express it in numbers, your knowledge is of a meagre and
unsatisfactory kind.}{William Thomson, Lord Kelvin, ``Electrical Units of
Measurement,'' lecture delivered at the Institution of Civil Engineers,
3 May 1883, in \emph{Popular Lectures and Addresses}, vol.~I, pp.~80--81
\cite[pp.~80--81]{hist:kelvin1891}.}

\chapmeta{Half-life: foundational. Archetypes: scaling (orders of
magnitude), failure (measurement error). Exercise target: 28. Project
track: household + analysis (hybrid). Hazard class: none. Reader path
default: standard, with sections explicitly tagged. Named cases: Mars
Climate Orbiter (1999), Air Canada Flight 143 (1983), Ariane 5 Flight 501
(1996), Cargill ground turkey Salmonella Heidelberg recall (2011).}

\noindent
We begin with measurement because every later engineering act depends on
a number that can be checked, repeated, and communicated. Before we
model, design, build, or repair, we measure.

\section{What measurement is for}\pathtag{core}

Measurement turns a private observation into a public engineering claim.
Once a quantity has been measured, it can be compared, reproduced,
communicated, archived, contracted, diagnosed, and calibrated. A
quantity that has not been measured can only be felt.

Older crafts carried quantities in trained bodies, tools, proportions,
marks, sounds, and repeated practice. The smith's judgement of iron, the
mason's judgement of stone, and the millwright's judgement of timber
could be precise in practice. What they could not always do was travel
cleanly through a drawing, contract, calibration certificate, or failure
investigation. A craft's knowledge often died with its last practitioner,
or survived in mute artifacts that the next generation had to
reverse-engineer.

Engineering does not have that luxury. A bridge in Pisa must take its
designed load whether the engineer who designed it is alive or dead. A
drug at $5 \,\si{\milli\gram}$ must be $5 \,\si{\milli\gram}$ whether
it is dispensed in S\~ao Paulo or Stockholm. A satellite in geostationary
orbit must speak to a ground station whose operator has never met its
builder. None of those is possible without an act that produces a number
which a stranger, on the other side of the world or a century later,
can reconstruct.

A recorded number is what lets the work leave the hands of its first
maker.

We can list the work measurement does:

\begin{itemize}[leftmargin=*]
  \item \engterm{Comparison}. Two beams, two batches, two designs, two
    suppliers. Measurement turns ``this one is better'' from a private
    impression into a public claim.
  \item \engterm{Reproduction}. A drawing without numbers can only be
    copied by the original artisan. With numbers, it can be made
    twice.
  \item \engterm{Communication}. Two engineers in different rooms work
    the same problem when they share a measurement; they argue past
    each other when they do not.
  \item \engterm{Memory}. A datasheet from 1972 still tells today's
    engineer what a thermostat does, because the datasheet's numbers
    carry units the engineer can read.
  \item \engterm{Contract}. When a manufacturer promises a tolerance,
    the customer can hold them to it because the tolerance is
    measurable.
  \item \engterm{Diagnosis}. A measurement that disagrees with
    prediction is a fault report. Without measurement, faults are
    mysteries.
  \item \engterm{Calibration}. A measurement is what a meter is
    \emph{for}, and a meter that is not calibrated is decoration.
\end{itemize}

Intuition can start each task; measurement is what lets another engineer
audit it.

The rest of Volume I turns this discipline into working habits: units,
calibration, traceability, error propagation, sensors, instruments,
statistics, and estimation. Volume II then gives those quantities
mathematical form, and Volumes III onwards apply the mathematics to the
world.

\begin{archetype}[Scaling]
Engineering decisions span enormous magnitudes. A single chapter may
need quantities ranging from the diameter of a transistor channel,
$\sim 10^{-9} \,\si{\meter}$, to the wingspan of a long-haul aircraft,
$\sim 10^{2} \,\si{\meter}$, to the orbit of a satellite,
$\sim 10^{7} \,\si{\meter}$. A working engineer carries an internal
ruler that maps any quantity in their domain to its order of magnitude
without conscious calculation. That internal ruler is built only by
repeated estimation. We will train it here, again in chapter 9 of this
volume, and continuously throughout the book.
\end{archetype}

\input{volumes/01-quantity/ch01-why-we-measure/figures/fig-si-dependency.tex}

\section{The impossibility of unmeasured engineering}\pathtag{core}

Consider a routine event: a foundation pour at a mid-sized construction
site. A truck arrives at 0830 with several cubic metres of ready-mix
concrete. The site engineer signs for it on a delivery slip that names
the mix design, the slump, the specified compressive strength, the
chloride content, the time of mixing, and the truck's identification.
Each entry is a number tied to a physical quantity. The slump comes
from a metal cone the carpenter knows; the strength is an MPa figure
that will be verified at 7 and 28 days from cylinders cast on site.

Without the numbers, the delivery slip no longer identifies an
engineering material. The truck could be carrying concrete that has sat
in the mixer for two hours and has lost its workability, a mix specified
for an inland slab dispatched to a coastal job where chloride ingress
will rust the rebar within a decade, a batch with the wrong cement type
for the soil's sulphate content, or the right concrete for the wrong
building. Each of those is concrete that someone judged to be good
without measuring it.

This pattern repeats in every engineering domain. A pharmacist preparing
a sterile dose without weighing the active ingredient cannot defend the
dose. A welder qualifying a joint without a procedure that records
voltage, current, travel speed, and shielding gas flow cannot reproduce
the weld. A network operator who reconfigures a router without recording
the change has chosen to forget what they did. In each case the work
may, by luck, be correct. None of it is engineering in the sense the
discipline requires.

Many engineering failures include a missing, ignored, or misread
measurement. We will study Tay Bridge, Tacoma Narrows, Bhopal,
Chernobyl, and Deepwater Horizon in Volume X only where the primary
record supports the mechanism. The lesson here is narrower: a decision
without a measurement record cannot be reconstructed after failure.

There are two senses in which an engineer can be said to ``know''
something. The first is the craftsman's knowledge: an internalised
sense, often correct, that this beam is sound or that bolt is loose.
The second is the engineer's knowledge: a recorded number, with units,
provenance, an instrument, a date, a witness, a calibration history,
and an error bar. The first kind of knowing is precious and must be
honoured. The second is what makes engineering possible at scale, in
public, and across time.

Throughout this work, engineering knowledge means the second kind.

\section{Estimation, error, precision, accuracy: the orienting
vocabulary}\pathtag{core}

Four words anchor measurement as a practice: estimation, error,
precision, accuracy. The reader who confuses any two of them will
write reports they cannot defend.

\subsection{Estimation}

\engterm{Estimation} is producing an answer to within a useful factor
before doing the calculation. The factor is usually three; sometimes
ten. The estimate is wrong, but it is wrong in an honest, recorded,
order-of-magnitude way. Engineers estimate before they calculate
because the calculation will produce a number whose plausibility the
estimate is the only check on. A spreadsheet with a unit error or a
buried minus sign can return a result that is wrong by a factor of
$10^{6}$. The estimate that preceded it would have warned.

We will use the following formal environment for estimation, and it
will recur in every chapter of every volume.

\begin{estimation}
\textbf{How heavy is a Boeing 747-400 at maximum take-off weight, in
metric tonnes?}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A working estimate goes through the following moves. A 747 carries
roughly 400 passengers; with crew and baggage, call that
$40 \,\si{\tonne}$. A wide-body airframe is several times heavier than
its passenger load; call the operating empty mass about
$200 \,\si{\tonne}$. Long-haul fuel is comparable to the empty-mass
scale; call it $150$ to $170 \,\si{\tonne}$. Add a small cargo and
reserve margin. Estimate: $390$ to $430 \,\si{\tonne}$.

Boeing lists the maximum 747-400 take-off weight as
$396{,}890 \,\si{\kilogram}$, or about $397 \,\si{\tonne}$
\cite{web:boeing-commercial-reference}. Our estimate brackets the
published value cleanly. The factor-of-three check passed. The
postmortem is that the empty-mass guess was the riskiest move; a
beginner's intuition can mistakenly anchor it on the payload, which
gives an impossible total. Sanity-checking each component against a
known bound (here, that empty mass cannot exceed maximum take-off
weight) is the habit the estimation environment exists to train.
\end{estimation}

The exercise trains the habit more than the answer. A reader who has
done five hundred such estimations by the end of the book has trained
an internal ruler that catches gross errors before the engineering
review catches them \cite{text:mahajan2014}.

\begin{estimation}
\textbf{What is the mass flow rate of global cement production, in
tonnes per second?}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A working estimate begins with the annual figure. Global cement
production is approximately $4 \times 10^9 \,\si{\tonne}$ per year
(current as of 2023). Convert to seconds: $1 \,\text{yr} \approx
3.15 \times 10^7 \,\si{\second}$. Therefore the mass flow rate is
$4 \times 10^9 / 3.15 \times 10^7 \approx 130 \,\si{\tonne\per
\second}$. The factor-of-three bracket is $40$ to $400 \,\si{\tonne
\per\second}$.

The USGS Mineral Commodity Summaries report $4.1 \times 10^9 \,\si{
\tonne}$ for 2023 \cite{data:usgs-mcs-cement-2024}. The mass flow
rate is therefore $4.1 \times 10^9 / 3.15 \times 10^7 \approx
130 \,\si{\tonne\per\second}$. The estimate matches the published
value to within a few percent.

The sanity check: a single large cement plant produces $\sim 1
\,\si{\tonne\per\second}$ at full capacity; the world operates
$\sim 130$ such plants in continuous production. The independent
estimate from the plant side agrees with the demand-side estimate
to within the same precision. The result implies the global
civil-and-construction engineering profession operates at a
continuous material-flow scale comparable to global steel ($\sim
60 \,\si{\tonne\per\second}$) and well above global aluminium
($\sim 2 \,\si{\tonne\per\second}$). The dataset
\texttt{data/cement\_production\_usgs.csv} (co-located with this
chapter) carries the year-by-year series from 2010 to 2023 for
readers who wish to repeat the calculation with current data.
\end{estimation}

A third worked estimation shows the habit protecting a calculation at
the exact point the chapter's accident cases failed: a unit conversion
crossing an interface.

\begin{estimation}
\textbf{A ground crew must load $22{,}300 \,\si{\kilogram}$ of jet fuel
onto an aircraft. The dispenser reads in litres and the fuel density is
$0.803 \,\si{\kilogram\per\liter}$. Estimate the volume to be
dispensed, and estimate the error if the crew instead uses the
imperial figure $1.77$ (pounds per litre) as though it were the metric
density.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The volume is mass over density. Working the order of magnitude first:
$22{,}300 \,\si{\kilogram}$ divided by roughly $0.8 \,\si{\kilogram\per
\liter}$ is about $2.8 \times 10^4 \,\si{\liter}$, near
$28{,}000 \,\si{\liter}$. Carrying the figures,
$V = 22{,}300 / 0.803 \approx 27{,}770 \,\si{\liter}$. The
factor-of-three bracket is $9{,}000$ to $83{,}000 \,\si{\liter}$; the
computed value sits comfortably inside it.

Now the wrong path. If $1.77$ is substituted for the density, the
crew computes $V = 22{,}300 / 1.77 \approx 12{,}600 \,\si{\liter}$,
which is $12{,}600 / 27{,}770 \approx 0.45$ of the correct volume. The
aircraft would depart with about $45\,\%$ of its intended fuel. The
error is not small and not random; it is the fixed factor
$1.77 / 0.803 \approx 2.2$, the pound-to-kilogram ratio, entering
where a density was required.

The estimation habit is the barrier. The order-of-magnitude check
$22{,}300 / 0.8 \approx 28{,}000 \,\si{\liter}$ takes a few seconds and
brackets the correct answer; the wrong path returns
$12{,}600 \,\si{\liter}$, which fails the bracket by more than a factor
of two and demands an explanation before a single litre is pumped. The
number $1.77$ carries units of pounds per litre, and pounds are not
kilograms; the dimensional check the estimate forces is exactly the
check the Gimli fuelling crew did not make. This is the same failure
narrated in section 1.4; here it is caught in advance by an estimate
that costs nothing.
\end{estimation}

\subsection{Error}

\engterm{Error} is the difference between a measured value and the
true value. In most engineering settings the true value is not known.
We estimate uncertainty from repeated measurements, instrument
resolution, calibration records, and propagation from input
quantities.

The international vocabulary of metrology defines measurement error as
a measured quantity value minus a reference quantity value, and
measurement uncertainty as a non-negative parameter characterising the
dispersion of quantity values attributed to the measurand
\cite{std:jcgm-vim}. The two are not the same: a measurement can have
a small error and a large uncertainty, or the reverse. We use the
words carefully throughout this work.

\subsection{Precision and accuracy}

\engterm{Precision} is the closeness of repeated measurements to one
another. \engterm{Accuracy} is the closeness of measurements to the
true value. They are independent in principle.

\begin{itemize}[leftmargin=*]
  \item Precise but inaccurate: a meter that consistently reads
    $5.32 \,\si{\volt}$ when the true voltage is $5.00 \,\si{\volt}$,
    with a spread of $\pm 0.01 \,\si{\volt}$. The numbers cluster
    tightly, but they cluster around the wrong value. The fix is
    calibration.
  \item Accurate but imprecise: a meter that averages
    $5.00 \,\si{\volt}$ across many readings but with a spread of
    $\pm 0.5 \,\si{\volt}$. Each reading is suspect; the average is
    correct. The fix is averaging more, or replacing the
    instrument.
  \item Both: a working laboratory standard.
  \item Neither: a display that drifts unpredictably and has not been
    calibrated.
\end{itemize}

A target diagram shows the four cases as four sets of shots on a
bullseye (\Cref{fig:vol01:ch01:target}). The distinction matters
because the corrective action differs. Precision is usually
improved by the instrument, environment, and procedure. Calibration
estimates the relation between indication and reference; adjustment
or correction is what improves the reported value in use.

\input{volumes/01-quantity/ch01-why-we-measure/figures/fig-target-precision-accuracy.tex}

\subsection{Type A and Type B evaluation}

The Guide to the Expression of Uncertainty in Measurement
\cite{std:iso-iec-guide98} distinguishes \engterm{Type A} uncertainty,
evaluated by statistical analysis of repeated measurements, from
\engterm{Type B} uncertainty, evaluated by other means: instrument
specifications, calibration certificates, manufacturer data,
documented experience. Both contribute to the combined standard
uncertainty, which is then expanded by a coverage factor (often
$k = 2$) to produce a reported expanded uncertainty.

Type A and Type B name how an uncertainty component is evaluated.
They do not mean random and systematic. Random and systematic
describe how errors behave in repeated measurements: a Type B
component can describe a systematic effect after correction, and a
Type A analysis can estimate repeatability under defined conditions.

The reader does not need the full machinery yet. The reader needs to
know that averaging can reduce the standard uncertainty associated
with independent random variation, but does not remove a shared bias.
For uncorrelated components combined in quadrature, the combined
standard uncertainty is at least as large as the largest component;
correlated components require the covariance term introduced in
chapter 4 of this volume and again, in propagation form, in chapter 5.
The Type A and Type B evaluation structure is summarised in
\Cref{fig:vol01:ch01:type-a-b}.

\input{volumes/01-quantity/ch01-why-we-measure/figures/fig-type-a-type-b.tex}

\subsection{The averaging floor: a worked combination}

The single most useful consequence of the Type A and Type B split is
quantitative: averaging drives the Type A component down and leaves
the Type B component untouched. A short worked combination fixes the
arithmetic before the reader meets the full propagation machinery in
chapter 4.

\begin{estimation}
\textbf{A pressure transducer has a repeatability of
$\sigma = 5 \,\si{\kilo\pascal}$ (Type A) and a
calibration-certificate bias uncertainty of
$u_B = 1.2 \,\si{\kilo\pascal}$ (Type B). How many averaged readings
does it take before the calibration uncertainty, not the noise,
dominates the reported value? What is the best combined standard
uncertainty the instrument can deliver, and how many readings buy
$90\,\%$ of that improvement?}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The Type A component of the mean is $u_A(n) = \sigma/\sqrt{n} =
5/\sqrt{n} \,\si{\kilo\pascal}$. The two components cross when
$u_A = u_B$, that is $5/\sqrt{n} = 1.2$, giving $\sqrt{n} = 4.17$ and
$n \approx 17$. Below seventeen readings the noise dominates; above,
the calibration bias dominates and further averaging is nearly wasted.

The best combined standard uncertainty is the floor the averaging
approaches: as $n \to \infty$, $u_A \to 0$ and
$u_c \to u_B = 1.2 \,\si{\kilo\pascal}$. No amount of averaging beats
$1.2 \,\si{\kilo\pascal}$ with this instrument and this certificate;
the only way past the floor is a better calibration.

The combined uncertainty is $u_c(n) = \sqrt{u_A(n)^2 + u_B^2} =
\sqrt{25/n + 1.44} \,\si{\kilo\pascal}$. At $n = 1$,
$u_c = \sqrt{26.44} = 5.14 \,\si{\kilo\pascal}$. Reaching within
$10\,\%$ of the floor means $u_c \le 1.1 \times 1.2 = 1.32$, so
$25/n + 1.44 \le 1.74$, giving $25/n \le 0.30$ and $n \ge 83$.
Eighty-three readings buy $90\,\%$ of the available improvement; the
next several hundred buy the last sliver. The engineer who logs a
thousand readings to chase the floor has misread where the
uncertainty lives. \Cref{fig:vol01:ch01:averaging-floor} plots the
two components and their quadrature sum; the flattening of the red
curve onto the amber floor is the visual form of this arithmetic.
\end{estimation}

\input{volumes/01-quantity/ch01-why-we-measure/figures/fig-averaging-floor.tex}

\begin{mastery}
The crossover $n^\ast = (\sigma / u_B)^2$ is worth carrying as a rule
of thumb: it is the number of averaged readings at which the random
component of the mean drops to the level of the fixed systematic
component. For the transducer above, $n^\ast = (5/1.2)^2 \approx 17$.
The rule exposes a design lever. If a program needs a combined
uncertainty below some target $u_{\text{target}}$ and the Type B
component already exceeds it, no averaging protocol can succeed; the
calibration must improve first. If the Type B component is comfortably
below $u_{\text{target}}$, the required sample size follows from
$u_A(n) = \sqrt{u_{\text{target}}^2 - u_B^2}$, so
\[
n \;=\; \frac{\sigma^2}{u_{\text{target}}^2 - u_B^2}.
\]
The denominator warns of the trap directly: as $u_B$ approaches
$u_{\text{target}}$, the required $n$ diverges, and the honest report
is that the target is unreachable with the present instrument rather
than reachable with heroic averaging. This is the same structure the
reader will meet again in chapter 4 as the variance-budget
allocation problem, where a fixed uncertainty target is apportioned
across independent contributors and the largest contributor sets the
achievable floor.
\end{mastery}

\subsection{Significant figures}

Significant figures are a coarse proxy for uncertainty. A reading of
$5.32 \,\si{\volt}$ asserts three significant figures, implying an
uncertainty around the third digit, roughly $\pm 0.005 \,\si{\volt}$.
The convention is sometimes sharper than the underlying measurement
supports and sometimes laxer. We use significant figures for quick
communication; we use an explicit uncertainty when the number
matters. A report with neither has not stated how much confidence its
number deserves. The three reporting conventions for the same
nominal value are compared in \Cref{fig:vol01:ch01:sig-figs}.

\input{volumes/01-quantity/ch01-why-we-measure/figures/fig-sig-figs.tex}

\subsection{Notation}

For the rest of this volume:

\begin{itemize}[leftmargin=*]
  \item A measured quantity is reported as $x \pm u(x)$, where $u(x)$
    is the standard uncertainty.
  \item An expanded uncertainty is reported as $x \pm U(x)$ with the
    coverage factor $k$ stated.
  \item A range is reported as $[x_{\min}, x_{\max}]$ with the
    coverage probability or rule for the range stated.
  \item Estimates that are explicitly within a factor (rather than
    additive) are reported as $\sim x$ with the factor stated.
\end{itemize}

The reader will see all four notations across the book. We do not
collapse them into a single house style because each carries
different information about what the writer believes.

\subsection{Resolution against tolerance: a worked match}

Precision, accuracy, and significant figures all describe what an
instrument reports. A separate question governs whether an instrument
may be used at all for a given task: is its resolution fine enough to
discriminate the tolerance the task allows? The two are matched at the
procedure-writing step, and a mismatch is a defect the reader can catch
with arithmetic alone.

\begin{estimation}
\textbf{A structural inspection calls for a fillet-weld throat of
$6.0 \,\si{\milli\meter}$ with a tolerance of
$\pm 0.5 \,\si{\milli\meter}$. Two instruments are on the bench: a
steel rule graduated in $1 \,\si{\milli\meter}$ and a weld-throat gauge
that reads to $0.1 \,\si{\milli\meter}$. Which may be used to certify
the throat, and what is the general rule?}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The reading resolution of an analog scale is about half a graduation,
so the steel rule discriminates to $\pm 0.5 \,\si{\milli\meter}$. The
tolerance band is $\pm 0.5 \,\si{\milli\meter}$ wide on each side, so
the rule's discrimination equals the full half-width of the allowed
window. An instrument whose reading uncertainty is as large as the
tolerance it must police cannot separate an in-tolerance throat from
one at the edge of failure; a $5.5 \,\si{\milli\meter}$ throat (out of
tolerance) and a $6.0 \,\si{\milli\meter}$ throat (nominal) can read
the same graduation. The rule fails the task.

The weld-throat gauge discriminates to $\pm 0.05 \,\si{\milli\meter}$
(half of its $0.1 \,\si{\milli\meter}$ graduation), a tenth of the
tolerance half-width. It resolves the $6.0 \pm 0.5 \,\si{\milli\meter}$
window into roughly twenty distinguishable steps and can certify the
throat with margin to spare. The gauge passes.

The general rule, sometimes stated as the ten-to-one or four-to-one
gauge-maker's ratio, is that the instrument's resolution (and, more
strictly, its total measurement uncertainty) should be a small
fraction of the tolerance it must verify, conventionally a quarter or
less, a tenth where the certification is critical. A resolution equal
to the tolerance is not marginally acceptable; it is disqualifying,
because the reading cannot distinguish the two sides of the boundary.
\Cref{fig:vol01:ch01:resolution-tolerance} shows the same relation for
a microlitre dosing task: the coarse instrument's reach spills past
both edges of the tolerance band, while the fine instrument sits inside
it.
\end{estimation}

\input{volumes/01-quantity/ch01-why-we-measure/figures/fig-resolution-tolerance.tex}

This is the arithmetic behind the medical-dosing exercise at the end of
the chapter and behind the coordinate-measuring-machine thermal clause
in the compliance audit later in it. The instrument is chosen against
the requirement, in writing, before the first reading; an instrument
selected because it was the one at hand is the quiet origin of a
measurement that cannot defend itself.

\section{Measurement crises: Mars Climate Orbiter, Air Canada Flight
143, Ariane 5}\pathtag{standard}

Three accidents show what happens when a number crosses an interface
without the protection it needs. We use abbreviated case summaries
here and keep the full seven-part case records in the named-cases
registry. Until that registry exists, these summaries are draft notes,
not release-ready case studies.

\subsection{Mars Climate Orbiter, 1999}

The Mars Climate Orbiter (MCO) was a NASA mission intended to
study the Martian climate from low Mars orbit, launched on
11 December 1998 from Cape Canaveral Air Station on a Delta II
7425 launch vehicle. It carried a Pressure-Modulator Infrared
Radiometer (PMIRR) for thermal-emission spectroscopy and a Mars
Color Imager (MARCI) wide-angle camera. Total mission cost was
\$$327.6$ million in $1998$ dollars (\$$125$ million spacecraft,
\$$93$ million launch, \$$109$ million ground operations and
science). On 23 September 1999 it executed the Mars Orbit
Insertion (MOI) burn at the wrong altitude, passed too close to
the planet, and was lost.

\paragraph{Technical mechanism.} The Mishap Investigation Board
(MIB) Phase I report, released on 10 November 1999 under chair
Arthur G. Stephenson (then director of NASA Marshall Space
Flight Center), traced the loss to a unit inconsistency at the
contractor-navigation interface \cite{acc:nasa-mco-mib}. The
SM\_FORCES ground software, developed by Lockheed Martin
Astronautics in Denver under contract NAS7-1407, computed
small-impulse performance data for the spacecraft's thruster
firings. Its AMD (angular-momentum-desaturation) output file was
required, by interface specification document JPL D-15110-1
(\textit{Mars Surveyor Operations Project: Mars Climate Orbiter
Mission Plan}), to contain impulse data in newton-seconds (the
SI unit per JPL convention), but Lockheed Martin's software
output the values in pound-force-seconds without the unit
conversion. The trajectory modellers at JPL, applying values to
the orbit-determination model verbatim, used small-forces values
low by a factor of $4.448$ (the lbf$_{s}$-to-N$_{s}$ conversion,
$1\,\text{lbf} = 4.448\,221\,615\,260\,5\,\si{\newton}$ exactly).

Each angular-momentum-desaturation manoeuvre applied a small
impulse the navigation model under-counted. Over the $9.5$-month
cruise from launch on 11 December 1998 to MOI on 23 September
1999, the cumulative trajectory error grew by a few kilometres
per AMD event. The MIB report identifies $10$ to $14$ AMD events
during cruise, each contributing an integrated
$\Delta v \approx 3\,\si{\milli\meter\per\second}$ error in the
navigation model. The trajectory error at Mars approach reached
the threshold where the closest-approach altitude was
approximately $57\,\si{\kilo\meter}$ rather than the planned
$226\,\si{\kilo\meter}$. The MOI burn nominally circularised the
orbit to a peri-areion altitude of $226\,\si{\kilo\meter}$ above
the Mars equator; at the actual $57\,\si{\kilo\meter}$
peri-areion, atmospheric dynamic pressure during the
high-velocity passage exceeded the spacecraft's structural
design margin. Telemetry was lost as planned at $T = 09{:}05{:}51$
PDT during planned signal occultation as the orbiter passed
behind Mars; signal was never re-acquired. The MOI burn lasted
$16$ minutes $23$ seconds; the spacecraft was destroyed at an
inferred time approximately $9$ minutes into the burn
\cite{acc:nasa-mco-mib}.

The accumulated bias was a sustained systematic error rather
than a sudden event. The trajectory team had identified
unexplained discrepancies in the months preceding the encounter
that they were unable to attribute to a specific source: the
inferred trajectory required a small additional $\Delta v$
adjustment at each navigation update to remain consistent with
the Earth-station Doppler tracking. The team hypothesised
gravitational anomalies, solar pressure mismodelling, and
gas-leak thrust, but the unit hypothesis was never raised in
recorded engineering communications. The MIB report concluded
that the cumulative pattern was identifiable in the residuals
six weeks before encounter, but the trajectory team's working
discipline did not include a systematic re-derivation of the
small-forces model against an independent unit-consistency
check.

\paragraph{Organisational context.} The failure sat in the interface
between contractor data and the navigation model: Lockheed Martin
Astronautics in Denver delivered the SM\_FORCES output in English
units; JPL's navigation model required newton-seconds per the
interface specification document. Multiple opportunities to detect
the discrepancy existed and were missed. The trajectory anomalies
during the cruise phase were investigated but the unit hypothesis
was not raised. The independent verification and validation
contractor was not staffed to perform end-to-end unit consistency
checks. The navigation team was understaffed and operating under
schedule pressure. The independent peer review process for the
trajectory analysis was not in place. The cumulative result was an
engineering organisation that had partitioned the technical work
across two contractors and two centres without a single party
accountable for the unit-consistency property across the partition.

\paragraph{Lesson for measurement.} The report's recommendations
named the missing barriers explicitly: verify consistent units
across design and operations, audit data transferred between JPL
and Lockheed Martin, verify the small-forces models against flight
data \cite{acc:nasa-mco-mib}. The deeper measurement lesson is the
chapter's central insistence: every quantity that crosses an
interface must carry its unit, and the conversion at the interface
must be a barrier (a checked, audited, automated transformation)
rather than a step (an unchecked memory operation in the receiving
party's processing chain). The interface specification is not
self-enforcing; the interface specification plus the conversion
barrier plus the periodic verification are the engineering
discipline. MCO was lost because the specification existed and the
barrier did not. \Cref{fig:vol01:ch01:mco-trajectory} sketches the
trajectory drift across the cruise.

\input{volumes/01-quantity/ch01-why-we-measure/figures/fig-mco-trajectory.tex}

\subsection{Air Canada Flight 143, 1983 (the Gimli Glider)}

Air Canada Flight 143, registration C-GAUN (a Boeing 767-233
delivered to Air Canada on 18 March 1983, the eighth $767$ off
the Renton production line), departed Montreal-Dorval at
$17{:}25$ Eastern Daylight Time on 23 July 1983 for a scheduled
Ottawa-Edmonton service with intermediate stop at Ottawa. With
$61$ passengers and $8$ crew, the aircraft ran out of fuel at
$41{,}000\,\text{ft}$ over Red Lake, Ontario at $20{:}11$ Central
Daylight Time and glided $89$ nautical miles to an unpowered
landing on runway 32L at the decommissioned Royal Canadian Air
Force station at Gimli, Manitoba. Captain Robert Pearson and
First Officer Maurice Quintal landed the aircraft at $20{:}38$
CDT. There were no fatalities; ten passengers were injured during
the emergency evacuation when the rear emergency slides did not
fully deploy due to the aircraft's nose-down attitude (the
collapsed nose gear after landing).

\paragraph{Technical mechanism.} The Board of Inquiry under
Justice George H. Lockwood reported on $26$ April $1985$
\cite{acc:lockwood-gimli1985}. The Boeing 767-200 was the first
Air Canada aircraft to use metric units throughout: fuel
quantity in kilograms on the flight deck, fuel load in kilograms
on the flight plan, payload in kilograms on the load sheet. The
fuelling process at the gate was still handled in litres
(Canadian commercial-aviation fuel deliveries used litres at the
truck, gallons at the storage facility). The aircraft's fuel
quantity processor channel (FQPC) computes the fuel mass from the
in-tank capacitance probes and the fuel density measured by an
in-tank densitometer; the C-GAUN's FQPC had failed during the
post-maintenance test at Edmonton on $5$ July $1983$. The
minimum-equipment-list (MEL) provision for an inoperative FQPC
required the use of manual drip-stick measurements at each fuel
loading, with the fuel mass computed from the drip-stick volume
multiplied by the fuel density from a separate measurement.

The fuelling crew at Montreal (the aircraft was refuelled before
the Ottawa-Edmonton leg) measured the dripstick volume in litres
and needed to compute the additional fuel required to bring the
total mass to the flight plan's $22{,}300\,\si{\kilo\gram}$ figure.
The dripstick reading indicated $7{,}682\,\si{\liter}$ already on
board. The fuel density was $0.803\,\si{\kilo\gram\per\liter}$ per
that day's refinery measurement. The required additional volume
to reach $22{,}300\,\si{\kilo\gram}$ total was therefore $(22300 -
7682 \times 0.803)/0.803 = (22300 - 6169)/0.803 = 20{,}088\,\si{
\liter}$.

The actual calculation used the imperial unit-derived conversion
the fuelling crew was familiar with: pound-mass per litre
$\approx 1.77\,\text{lb/L}$ for jet fuel at the same density.
Substituting $1.77$ where $0.803$ was required, the crew
computed the additional load to be $(22300 - 7682 \times 1.77)/
1.77 = (22300 - 13597)/1.77 = 4{,}917\,\si{\liter}$. The crew
delivered $4{,}917\,\si{\liter}$ rather than $20{,}088\,\si{
\liter}$, leaving the aircraft with a total of $12{,}599\,\si{
\liter}$ of fuel rather than the planned $27{,}770\,\si{\liter}$
($22{,}300\,\si{\kilo\gram}$). The fuel load on board was
approximately $45\%$ of the flight plan's requirement.

The aircraft departed Montreal with sufficient fuel to reach
Ottawa and a portion of the way toward Edmonton. At
$20{:}10$ CDT the left engine flamed out at $41{,}000\,\text{ft}$
from fuel exhaustion; the right engine flamed out approximately
six minutes later. The 767's ram air turbine (RAT, a small wind-
driven propeller that deploys from the underside of the
fuselage) deployed automatically and provided hydraulic pressure
to the flight controls but not to the landing gear. Captain
Pearson, a former glider pilot, executed a forward-slip
manoeuvre (a technique unfamiliar to most airline pilots) to
increase the descent angle and bleed off altitude during final
approach to runway 32L at Gimli. The nose gear failed to extend
fully through gravity drop; the aircraft landed on the main gear
and slid on the nose with the unextended nose gear collapsed.
The runway at Gimli, designated 32L, had been converted to a
drag strip; spectators and racers were on and near the runway
during the landing. There were no fatalities to occupants or
ground personnel \cite{acc:lockwood-gimli1985}.

\paragraph{Organisational context.} The unit-system confusion was
contributory rather than singular. The 767's introduction had
included kilograms in the load calculation workflow but had not
withdrawn the imperial conversion factors that the rest of the
Air Canada fleet still used; the fuelling crew worked the calculation
using the conversion they were familiar with. The unserviceable
fuel quantity processor had been authorised for dispatch under
minimum-equipment-list provisions that permitted manual drip-stick
calibration. The dispatch chain trusted the fuelling crew's
arithmetic without independent verification. The aircraft was
loaded against a fuel-mass calculation that no independent party
had checked against the flight plan's fuel requirement. Each of
these is a real lesson in itself; we take them up in Volume X
chapter 11. The Board's recommendations included specific changes
to the load-calculation procedure, the dispatch authority, and the
unit-system transition management.

\paragraph{Lesson for measurement.} The measurement lesson is the
narrow one: when two unit systems pass through the same procedure,
the conversion must be a barrier, not a step. The fuel-mass
calculation crossed unit systems three times (volume in litres,
density in $\text{lb/L}$ or $\si{\kilogram\per\liter}$, mass in
kilograms for the load) without an interface that flagged the
inconsistency. The seven-component measurement habit of section
1.5 (with the units explicit at every step and a second engineer's
cross-check on every conversion) would have caught the error at the
gate. The safe landing does not vindicate the fuelling system; it
shows that crew skill recovered after the measurement system had
already failed.

\subsection{Ariane 5 Flight 501, 1996}

On 4 June 1996 the maiden flight of the Ariane 5 launcher
(Flight 501), launch vehicle serial number AR501, launched from
the European Spaceport at Kourou, French Guiana at $12{:}33{:}59$
UTC carrying the four Cluster scientific satellites (Cluster
SAT-1 through SAT-4, magnetospheric research spacecraft developed
by ESA). The launch vehicle began to disintegrate at $T+39\,\si{
\second}$ after main engine ignition ($T+31\,\si{\second}$ after
lift-off, lift-off occurring at $T+8\,\si{\second}$ after engine
ignition for vehicle clearance). The inquiry board chaired by
Jacques-Louis Lions of the Acad\'emie des sciences reported on
$19$ July $1996$ \cite{acc:ariane501-ifr}, $45$ days after the
failure; the report is among the fastest substantive accident
inquiries in space history.

\paragraph{Technical mechanism.} The cause was a software
exception in the Syst\`eme de R\'ef\'erence Inertielle (SRI), the
inertial-reference platform that delivers attitude and velocity
data to the On-Board Computer (OBC). The exception was triggered
by a $64$-bit floating-point representation of the BH variable
(\textit{Biais Horizontal}, horizontal-bias term, the integrated
horizontal velocity used to compensate for sensor misalignment)
being converted to a $16$-bit signed integer using an Ada
language conversion path. The value of BH at the moment of the
conversion was approximately $93\,000$ in the relevant internal
units (decimetres-per-second after scaling); the maximum
representable value in $16$-bit signed integer arithmetic is
$32\,767$. The conversion overflowed, raising the Ada predefined
exception \texttt{Constraint\_Error}.

The conversion routine was inside the SRI's alignment function
(\textit{SRI Alignement Fin}, the precision-alignment computation
that continued running for $50\,\si{\second}$ after lift-off in
the Ariane 4 SRI software inherited by Ariane 5). The Ariane 4
flight envelope kept BH well below $32\,767$ in all reachable
trajectories; the Ariane 4 qualification did not need to consider
the protection of this conversion. The Ariane 5 vehicle's larger
mass and different ascent profile produced horizontal velocity
that reached the overflow value approximately $36\,\si{\second}$
into flight. Ariane 5 had no need for the SRI alignment function
to continue running after lift-off, but the software had been
inherited from Ariane 4 without modification on the principle
that re-qualified software costs more than software-as-is reuse.

Of the seven operand conversions in the alignment function, four
were protected by the Ada \texttt{...range...} clause; three were
not, including the BH conversion. The lack of protection was a
documented engineering decision in the Ariane 4 software
qualification: the three unprotected conversions had been
demonstrated safe by the Ariane 4 flight envelope. Their
unprotected status was carried to the Ariane 5 software without
revisiting the demonstration against the Ariane 5 flight envelope.
The exception was raised, caught by the Ada runtime as
\texttt{Constraint\_Error}, and the SRI shut down at
$T+36.7\,\si{\second}$. The back-up SRI, executing the same
software, encountered the same exception approximately $72\,\si{
\milli\second}$ later and shut down at $T+36.77\,\si{\second}$.

With both inertial-reference systems shut down, the launch
vehicle's main computer received diagnostic bit patterns where
it expected attitude data and interpreted them as nominal-but-
extreme angular-velocity and attitude values. The OBC commanded
correspondingly extreme thrust-vector deflections (full deflection
of the solid-rocket-booster nozzles and the cryogenic main-stage
nozzle); the aerodynamic loads at the trajectory's high dynamic
pressure exceeded the structural margin; the vehicle began to
break up at $T+39\,\si{\second}$. The Range Safety Officer
commanded the self-destruct system at $T+40\,\si{\second}$, and
the destruction sequence completed by $T+42\,\si{\second}$.
The four Cluster scientific satellites and their associated
launch costs, valued at approximately \$$370$ million in 1996
US dollars (the Cluster mission was eventually re-flown as
Cluster II in 2000 on four Soyuz-Fregat launches at a cost of
\$$120$ million per quartet), were destroyed.

\paragraph{Organisational context.} The exception occurred inside
the SRI alignment function, which was still running after lift-off
because of an Ariane 4 requirement and was not needed for Ariane 5
after lift-off. The Ariane 5 qualification work had not adequately
analysed or tested that SRI software against the Ariane 5 flight
domain; the horizontal bias value reached in the first seconds of
Ariane 5 flight was approximately $5 \times$ the maximum value the
Ariane 4 trajectory ever produced. The decision to reuse the
qualified Ariane 4 SRI software without re-qualification was
explicit and documented; the analysis supporting the decision
focused on the SRI hardware behaviour and did not extend to the
software's protected-conversion paths. The verification regime
for software reuse across launcher families was subsequently
revised; the Lions Report's recommendations included specific
provisions for envelope-based requalification of reused software,
for protected type-conversion practices, for dissimilar backup
software, and for the development of operational simulators
covering the full flight envelope of the new launcher.

\paragraph{Lesson for measurement.} The missing check was not a
sensor reading in flight; it was a verification test of the
reused software against the new trajectory. The measurement
framing of the failure is the range-and-validation-envelope
discipline: every measured quantity carries an expected range,
and every consumer of that quantity must have its behaviour
verified against the full range the producer can deliver in the
deployment environment. The horizontal-bias variable was a
``measurement'' in the sense that it was a computed quantity
intended to represent a physical state of the vehicle; its range
in Ariane 5 deployment exceeded its range in Ariane 4 deployment.
The reuse-without-requalification was the engineering decision
that propagated the un-checked-range condition into the new
deployment. Volume X chapter 9 returns to this failure mode as
``silent confusion across an interface''; the chapter's measurement
lesson is the narrower one: every number that crosses an interface
must carry its range, its unit, and its validation envelope, and
the receiving system must enforce all three.

\subsection{The pattern}

Mars and Gimli are unit-interface failures: two systems with
different number conventions met at an interface, and the conversion
was treated as memory work rather than as a barrier. Ariane 501 is a
range and reuse-envelope failure: software qualified for one vehicle
was reused on another whose flight domain exceeded the data type's
range. The shared discipline is interface measurement: units, ranges,
data types, provenance, and validation envelope must be explicit
before integration.

Volume X returns to this family under the name \emph{silent confusion
across an interface}, with a longer list. The lesson here is basic:
every number that crosses an interface must carry its unit, range,
and provenance. Without those, the receiving system cannot know what
quantity it has received.

\subsection{Long-form case: the Ariane 501 forensic walk}\pathtag{standard}

The three preceding cases each carry a registry entry; the chapter
narrative compresses each to its lesson. We here walk one of them at
the level of the post-flight inquiry's reconstruction, to make visible
the chain of decisions that turn a single representational mismatch
into a vehicle loss. The case is Ariane 501. The inquiry report is
\cite{acc:ariane501-ifr}.

\paragraph{Vehicle and mission.} Ariane 5 was the European Space
Agency's heavy-lift launcher developed during the late 1980s and
early 1990s to replace Ariane 4. The maiden flight, designated
Ariane 501, lifted off from Kourou on $4$ June $1996$ carrying the
Cluster scientific payload. The flight ended at $T+39
\,\si{\second}$ when the launcher self-destructed at an altitude of
approximately $4000 \,\si{\meter}$. The Cluster mission was lost.
The financial cost to the European Space Agency was approximately
\$370 million for the vehicle and another \$500 million for the
satellites.

\paragraph{The reused subsystem.} The Inertial Reference System (SRI)
was a subsystem of the launcher's guidance and control architecture.
Two redundant SRIs ran in parallel; each performed inertial
integration of accelerations and angular rates from a three-axis
gyroscope-and-accelerometer cluster. The SRI hardware and software
were inherited from Ariane 4 with minimal modification. The
inheritance decision was a deliberate engineering economy: the
Ariane 4 SRI was field-proven over more than one hundred flights;
re-developing the SRI for Ariane 5 would have cost approximately
\$50 million and added two years to the development schedule.

\paragraph{The alignment routine.} Inside the SRI, a routine
performed an alignment calculation for the inertial platform. The
routine had been written for Ariane 4 to support a specific
pre-launch procedure: it allowed the launch to be paused and resumed
within a window of $50 \,\si{\second}$ after liftoff, in case of a
late hold. The procedure was a peculiarity of the Ariane 4 launch
campaign and was not needed on Ariane 5; the latter's launch
procedure did not allow a hold-and-resume of this kind. The
alignment routine was nevertheless retained in the inherited
software, set to run for $50 \,\si{\second}$ after liftoff on every
flight as a legacy.

\paragraph{The conversion.} Within the alignment routine, a 64-bit
floating-point value representing the horizontal velocity component
was converted to a 16-bit signed integer. The conversion was
performed without a range check; the assumption embedded in the
Ariane 4 code was that the value would remain within the
representable range of $-2^{15}$ to $+2^{15} - 1$ during the
$50 \,\si{\second}$ window. On Ariane 4, the assumption was correct:
the horizontal velocity at $T+50 \,\si{\second}$ was bounded below
the integer-overflow threshold by the launcher's flight profile.

\paragraph{The new envelope.} Ariane 5 had a different
trajectory. Its booster's higher thrust-to-weight ratio and its
pitch-over profile gave it higher horizontal velocity earlier in
flight. At approximately $T+37 \,\si{\second}$, the horizontal
velocity exceeded the 16-bit signed integer's representable range.
The conversion in the alignment routine raised an operand-error
exception.

\paragraph{The exception handler.} The SRI's exception-handling
software treated the operand error as a hardware failure of the
inertial platform. The handler followed the system's default
behaviour: it sent a diagnostic word to the on-board computer,
recorded the failure in non-volatile memory, and shut down the SRI.
The exception fired in the active SRI at $T+37.2 \,\si{\second}$;
the backup SRI took over.

\paragraph{The redundancy fallacy.} The backup SRI was running the
same software, performing the same alignment calculation, with the
same range mismatch. The backup SRI's conversion raised the same
exception approximately $72 \,\si{\milli\second}$ after the primary
SRI's shutdown. The backup SRI also shut down. The launcher had no
inertial reference data.

\paragraph{The cascade.} The on-board computer interpreted the
diagnostic word from the active SRI (which contained the SRI's
exception trace, not a flight-state vector) as flight data. The
computer used the bit pattern as if it represented an angle of
attack. The interpreted angle of attack was nonsense and indicated a
large nose-up pitch; the guidance system commanded a full nozzle
deflection of the boosters to correct. The aerodynamic loads
produced by the nozzle deflection exceeded the structural design
limits of the booster-to-core attachment. The booster began to
separate from the core stage. The Range Safety Officer's automatic
self-destruct system, sensing the breakup, fired the explosive
charges. The vehicle was destroyed at $T+39 \,\si{\second}$.

\paragraph{The cost layers.} The direct cost was the vehicle and the
satellites, approximately \$870 million. The indirect cost included
a one-year delay in the Cluster programme, a two-year delay in
commercial Ariane 5 service entry, and a re-certification campaign
for the launcher that consumed approximately \$300 million in
additional development. The reputation cost to the European Space
Agency and to Arianespace, the commercial operator, was substantial
and is the subject of internal documentation we do not narrate
further.

\paragraph{The decision points.} The inquiry identified the
following sequence of engineering decisions that, taken together,
made the failure inevitable.

(i) The inheritance of the SRI software from Ariane 4 without a
flight-envelope qualification against Ariane 5's trajectory. The
inquiry's analysis is the closest the report comes to identifying a
root cause; the report's language is more careful, but the chain of
recommendations targets the qualification gap directly.

(ii) The retention of the alignment routine that was not needed on
Ariane 5. The routine was a known legacy at the time of
integration; the decision to retain it was made on cost grounds and
on a judgement that ``inherited software is more reliable than new
software,'' a judgement that holds only when the inheritance
envelope is unchanged.

(iii) The absence of a range check on the conversion. This decision
was inherited from Ariane 4 and was a defensible omission on that
vehicle, because the range was bounded by the flight profile. On
Ariane 5, the inherited absence was a defect; the qualification
campaign should have flagged it.

(iv) The redundant SRI architecture's reliance on identical
software. The architecture was designed against hardware faults,
which independent units would not share. It was not designed
against software faults, which identical units would share. The
inquiry recommended dissimilar-software redundancy for future
mission-critical reusable subsystems, a recommendation now
incorporated into ESA's mission-assurance practice.

(v) The exception handler's default behaviour. Treating the operand
error as a hardware failure and shutting down the SRI was the wrong
response to a software defect; the handler should have masked the
failed conversion, used the last valid integer value, and continued
the alignment calculation degraded rather than terminated. The
post-Ariane 501 corrective added range checks and graceful-
degradation paths to the SRI's exception handler.

\paragraph{The chapter's lesson.} The Ariane 501 case is sometimes
narrated as a software bug. The narrative compresses the failure
into a single line of code: the unprotected type conversion. The
single line is real, and the corrective at the software layer is
real. The deeper lesson is upstream of the software: the inherited
subsystem carried its qualification envelope with it, and the
re-qualification step that should have re-validated the envelope
against the new vehicle was not performed. The single line of
code is the place where the envelope mismatch surfaced; the
qualification process is the engineering layer where the mismatch
should have been caught.

The lesson generalises beyond Ariane 5. Every reused subsystem
inherits its envelope. Every new vehicle, new mission, or new
operating condition is an envelope change. The verification step
that re-establishes the envelope match is the discipline this
chapter has named in compressed form; the Ariane 501 walk-through
is the discipline at the scale of an actual launcher loss.

\section{The measurement habit}\pathtag{core}

A measurement is not a single number. A measurement is a number
together with its date, its instrument, its operator, its conditions,
its calibration history, and its provenance. Strip any of those and
the measurement loses what makes it engineering. We name the habit
explicitly here, because every later chapter assumes the reader has
acquired it.

The measurement habit has seven concrete components.

\subsection{The dated entry}

Every measurement is recorded with the date and, when relevant, the
time. A figure dated only by the year ages silently. A reading taken
at $14{:}30$ on a hot afternoon at a steel mill is not the same
reading as one taken at $05{:}30$ in winter; if a chapter cites a
reading without a time, the reader cannot judge whether the
ambient was relevant.

\subsection{The instrument and its calibration}

A measurement carries the name and serial number of the instrument
that produced it, and the calibration history of that instrument: who
calibrated it, when, against which reference, and to what tolerance.
A measurement made with an uncalibrated instrument is, formally, a
reading rather than a measurement. Calibration is the subject of
Chapter 3.

\subsection{The operator}

A measurement carries the name of the operator. This is sometimes
read as bureaucratic; it is, in fact, a defence against silent
mistakes. When a reading is contested, the engineer who signs the
log can be asked what they did. An anonymous reading cannot be
defended.

\subsection{The conditions}

Temperature, humidity, vibration, line voltage, ambient illumination,
elapsed time since the instrument was warmed up, posture of the
sample. Most measurements depend on at least one of these. The
working engineer learns by experience which conditions matter for
which measurement; in the meantime, the safe rule is to record the
ones that plausibly could.

\subsection{Provenance of the reference}

If a measurement uses a reference (a calibration mass, a standard
voltage source, a reference thermometer, a reference clock), the
provenance of that reference is part of the measurement. The
traceability chain is the subject of chapter 3 of this volume.

\subsection{The reading itself}

The reading is recorded with the appropriate number of significant
figures. Where a digital instrument's display flickers between two
values, both are recorded, or the average and the spread. A reading
written with one digit beyond the instrument's resolution is a
fiction.

\subsection{The estimate that preceded it}

Where the measurement is an exercise in checking a calculation, the
reader's estimate is recorded \emph{before} the measurement. Otherwise
the human tendency is to revise the estimate to match the result. The
recorded estimate is a defence against the engineer's own confirmation
bias.

A laboratory or engineering notebook in which these seven appear
routinely is what we will call, throughout this work, a \engterm{good
record}. The first project in this chapter is to keep a notebook of
that kind for one week.

\subsection{Five worked notebook entries}\pathtag{standard}

The schema above is named in compressed form, and a reader who has
never kept a calibrated-instrument log can read it without knowing
what populating each field looks like. The five entries below show
the schema worked through five very different measurements: a
mechanical reading on a structure, a chemical reading at the bench, a
thermal reading on a circuit board, a frequency reading on a power
line, and a slow mass-loss reading on an evaporating sample. Each
entry contains every schema field. The reader's own notebook entries
during the project at the end of this chapter should look like one of
these.

\paragraph{Entry 1: structural deflection logged over a week.}

\begin{description}[leftmargin=2em, itemsep=2pt]
  \item[Date / time.] 2026-04-13 (Monday) through 2026-04-19
    (Sunday), readings at 08{:}30 and 16{:}30 each weekday plus a
    single 12{:}00 reading on Saturday and Sunday.
  \item[Instrument.] Mitutoyo dial indicator series 2046S-08,
    $10 \,\si{\milli\meter}$ range, $0.01 \,\si{\milli\meter}$
    graduation, serial no.\ 4127, indication accuracy
    $\pm 0.013 \,\si{\milli\meter}$ over full range, retrace error
    $0.01 \,\si{\milli\meter}$ \cite{web:ch01t3-mitutoyo-2046S}. Last
    calibration: 2025-09-15 (A2LA-accredited calibration laboratory,
    cert.\ no.\ MIT-2046S-08-4127-250915, scope 5--250
    \si{\milli\meter} length). Calibration due: 2026-09-15.
  \item[Measurand and setup.] Mid-span vertical deflection of a
    simply-supported ASTM A572 grade 50 steel I-beam, W6$\times$12
    section, span $L = 2.000 \,\si{\meter}$, under a point load of
    $W = 50 \,\si{\newton}$ applied at mid-span by a calibrated
    deadweight (Troemner class F mass set, cert.\ no.\
    TR-F-2024-118).
  \item[Operator.] J.~R.~(student lab tech; initials in notebook
    margin; full name on a signed front-leaf inventory).
  \item[Conditions.] Lab ambient logged each reading session:
    week mean $T = 21.4 \pm 0.6 \,\degC$, RH
    $42 \pm 5 \,\si{\percent}$, line voltage $120.1 \,\si{\volt}$
    (uncontrolled but irrelevant for this measurand), ambient floor
    vibration below the gauge's noise floor (verified by a 60-s
    null reading with the load removed).
  \item[Provenance of reference.] Length scale of the dial gauge
    traceable to NIST through the A2LA chain: NIST gauge-block
    primary $\to$ A2LA accredited lab transfer standards $\to$
    Mitutoyo factory calibration $\to$ the cert.\ on file. Load mass
    traceable to NIST through the Troemner F-class deadweight chain.
  \item[Estimate before reading.] First-principles beam deflection
    for a centrally-loaded simply-supported beam,
    $\delta = W L^3 / (48 E I)$, with $E = 200 \,\si{\giga\pascal}$
    and $I = 9.1 \,\si{\centi\meter}^4$ for W6$\times$12 about its
    weak axis gives $\delta \approx 50 \cdot 2.000^3 /
    (48 \cdot 200 \times 10^9 \cdot 9.1 \times 10^{-8}) \approx
    0.46 \,\si{\milli\meter}$. Recorded in the notebook 2026-04-12
    before any reading was taken.
  \item[Reading.] Twelve readings, mean
    $\bar\delta = 0.47 \,\si{\milli\meter}$, sample standard
    deviation $s = 0.012 \,\si{\milli\meter}$, drift across the week
    $+0.02 \,\si{\milli\meter}$ between Monday morning and Friday
    afternoon (consistent with the daytime thermal expansion of the
    support pillars; recorded as such).
  \item[Signature.] J.~R., dated 2026-04-19; counter-signed by the
    lab supervisor on 2026-04-22.
\end{description}

The entry occupies a single facing page. The estimate-then-measure
discipline put a number on the page before the gauge moved; the
agreement with that estimate to within $1\sigma$ is what lets the
reader believe the gauge was reading the structure rather than
something else.

\paragraph{Entry 2: chemical titration with reagent traceability.}

\begin{description}[leftmargin=2em, itemsep=2pt]
  \item[Date / time.] 2026-03-04, 10{:}15.
  \item[Instrument.] Class A graduated burette, 50 \si{\milli\liter}
    capacity, tolerance $\pm 0.05 \,\si{\milli\liter}$ per ASTM E287
    \cite{std:ch01t3-astm-e287-2018}, serial no.\ BG-50A-2024-117. End
    point detected visually against a phenolphthalein indicator and
    confirmed by a pH probe (Mettler-Toledo InLab Routine Pro,
    cal.\ to pH 4.01, 7.00, 10.01 buffers on the day; cert.\
    MT-IRP-2603).
  \item[Measurand and setup.] Concentration of an unknown HCl
    solution, expected near $0.10 \,\si{\mol\per\liter}$, titrated
    with $0.1000 \,\si{\mol\per\liter}$ NaOH standardised the
    previous day against NIST SRM 84n potassium hydrogen phthalate
    \cite{std:ch01t3-nist-srm-84}.
  \item[Operator.] M.~A.~D.
  \item[Conditions.] $T = 22.7 \,\degC$, RH
    $38 \,\si{\percent}$, fume hood draught steady at
    $0.45 \,\si{\meter\per\second}$ face velocity, sample volume
    $25.00 \,\si{\milli\liter}$ measured by Class A pipette.
  \item[Provenance of reference.] NaOH titrant: J.~T.~Baker reagent
    SR-9469, lot 0089234, certificate of analysis dated 2025-11-12,
    expiry 2027-03. Standardisation tied to NIST SRM 84n through a
    1.5-month-old in-house ladder (recorded on the previous page);
    the SRM 84n certificate carries the NIST Inorganic Chemistry
    Division stamp and an expanded uncertainty of $0.05 \,\si{\percent}$
    (relative, $k=2$).
  \item[Estimate before reading.] An expected endpoint near
    $25.0 \,\si{\milli\liter}$ if the unknown is at the nominal
    $0.10 \,\si{\mol\per\liter}$. Recorded.
  \item[Reading.] Three replicate titrations, endpoints
    $25.07, 25.10, 25.04 \,\si{\milli\liter}$. Mean
    $25.07 \,\si{\milli\liter}$, sample standard deviation
    $0.030 \,\si{\milli\liter}$, implied unknown concentration
    $0.1003 \,\si{\mol\per\liter}$, $\pm 0.0004 \,\si{\mol\per\liter}$
    type-A on the endpoint plus $\pm 0.0001 \,\si{\mol\per\liter}$
    type-B from the NaOH standardisation.
  \item[Signature.] M.~A.~D., 2026-03-04; verified by lab manager
    A.~K., 2026-03-05.
\end{description}

The chain from the burette through the NaOH lot back to NIST SRM 84n
is what lets a stranger replicate the determination. The
three-replicate endpoint spread is the immediate evidence the reader
can show that the method is in control on the day.

\paragraph{Entry 3: circuit-board temperature swept across operating
conditions.}

\begin{description}[leftmargin=2em, itemsep=2pt]
  \item[Date / time.] 2026-02-19, 09{:}00 through 17{:}00 (six
    operating-condition sweeps).
  \item[Instrument.] Four Omega TC-K-FF-SLE-72 ungrounded type-K
    thermocouples \cite{web:ch01t3-omega-tck} (special limits of
    error, $\pm 1.1 \,\degC$ over $0 \,\degC$ to $+1250 \,\degC$
    per ASTM E230) bonded with Omegabond 200 thermally conductive
    epoxy at the four hottest spots of an ATCA carrier board
    identified by an FLIR T620 thermal-imaging pre-survey. Reading
    head: Omega RDXL4SD four-channel data logger with built-in
    cold-junction compensation, serial no.\ RDX-22-4187,
    last cal.\ 2025-12-08 (UKAS-accredited, scope $-200 \,\degC$ to
    $+1372 \,\degC$, certificate UK-RDXL4SD-2512-08-4187).
  \item[Measurand and setup.] Steady-state surface temperature at
    each of four hotspots (CPU package top, NIC ASIC die-attach
    pad, point-of-load converter inductor case, DRAM module DIMM
    edge) under six operating conditions: idle, 50\% load, 100\%
    load, each at chamber ambients $25 \,\degC$ and $45 \,\degC$.
    Forced-air cooling at $3.5 \,\si{\meter\per\second}$ across
    the board, measured by a TSI 9545 anemometer.
  \item[Operator.] B.~S.\ (electrical engineer, thermal team).
  \item[Conditions.] Chamber: Espec SH-242 environmental chamber,
    last cal.\ 2025-08-30, $\pm 0.5 \,\degC$ uniformity at setpoint.
    Power supply: HP 6675A in current-limit mode at 12.0 V, 22 A
    at full load. Line voltage stable at $120.1 \,\si{\volt}$. Each
    sweep allowed a $20 \,\si{\minute}$ soak before logging.
  \item[Provenance of reference.] ITS-90 temperature scale
    \cite{web:bipm-cct-its90} realised through the UKAS-accredited
    chain: NPL SPRT primary $\to$ UKAS lab transfer standard $\to$
    Omega factory cal.\ of the RDXL4SD against an SPRT-traceable
    triple-point-of-water cell and a tin freeze-point cell.
  \item[Estimate before reading.] Datasheet-derived junction-to-case
    rise for the CPU package at 100\% load and the rated thermal
    resistance gives a predicted case temperature of $\sim 78
    \,\degC$ at $25 \,\degC$ ambient and $\sim 98 \,\degC$ at
    $45 \,\degC$ ambient. Recorded on the cover sheet of the sweep.
  \item[Reading.] Twenty-four readings (four channels $\times$ six
    sweeps); the CPU case at the worst corner read
    $96.3 \pm 0.4 \,\degC$ (mean over the last 5 minutes of soak,
    type-A only). All channels within $\pm 1.5 \,\degC$ of estimate
    except the inductor case, which read $11 \,\degC$ hotter than
    estimate at full load (the inductor saturation-current
    derating, missed in the estimate, was the cause).
  \item[Signature.] B.~S., 2026-02-19; reviewed by hardware lead
    J.~F., 2026-02-20.
\end{description}

The inductor disagreement is the reason the estimate column exists.
A logged measurement that agrees with estimate to within the
combined uncertainty is uninteresting evidence the instrument is
working; a logged measurement that disagrees is a fault report on
either the model, the part, or the rig, and the engineer is forced
to settle which.

\paragraph{Entry 4: mains-frequency drift over 24 hours.}

\begin{description}[leftmargin=2em, itemsep=2pt]
  \item[Date / time.] 2026-01-22, 18{:}00 (start) through 2026-01-23,
    18{:}00 (end).
  \item[Instrument.] Keysight 53230A 350~MHz universal frequency
    counter \cite{web:ch01t3-keysight-53230A}, serial no.\
    US53120000235, internal high-stability OCXO timebase upgraded to
    GPS-disciplined operation through a Trimble Thunderbolt E
    receiver (Trimble Sunnyvale; serial no.\ TBE-08-14933).
    Last cal.\ 2025-11-02 (NIST-traceable, A2LA-accredited
    calibration laboratory, certificate KS-53230A-235-251102).
    Counter frequency accuracy with GPS lock: $5 \times 10^{-12}$
    over 24 h.
  \item[Measurand and setup.] North American Eastern Interconnection
    mains frequency, sampled from a 60 \si{\volt} step-down isolation
    transformer (Stancor P-8625) feeding the counter through a
    low-pass conditioning filter. Counter gate time 1 s, logging at
    $1 \,\si{\hertz}$ to an SD card.
  \item[Operator.] L.~M.~(power-systems student, undergraduate
    senior project).
  \item[Conditions.] Lab $T = 20.8 \,\degC$, RH
    $35 \,\si{\percent}$, no peak-shaving events in the load
    forecast (verified against the regional ISO day-ahead schedule).
  \item[Provenance of reference.] Counter timebase locked to GPS L1
    C/A code, traceable to USNO master clock and through it to a
    caesium-fountain primary $\to$ TAI, with an Allan deviation
    below $1 \times 10^{-12}$ at $\tau = 1000 \,\si{\second}$. Mains
    frequency nominal $60.000 \,\si{\hertz}$
    \cite{web:ch01t3-nerc-mains-frequency}.
  \item[Estimate before reading.] Expected mean within
    $\pm 2 \,\si{\milli\hertz}$ of nominal over 24 h under normal
    grid operation; expected $1\sigma$ short-term deviation of
    $\sim 10 \,\si{\milli\hertz}$. Recorded.
  \item[Reading.] 86\,400 one-second samples; mean
    $\bar f = 59.9982 \,\si{\hertz}$, sample standard deviation
    $9.4 \,\si{\milli\hertz}$, time error accumulated over 24 h
    $-2.6 \,\si{\second}$ (consistent with a small chronic
    under-correction of the integrated time error). Maximum
    excursion $+38 \,\si{\milli\hertz}$ during the morning-rise
    transient at 06{:}40--07{:}10.
  \item[Signature.] L.~M., 2026-01-23; faculty advisor C.~B.,
    2026-01-26.
\end{description}

A 1 ppm grid measurement of any duration would have been a research
artefact in 1950. The GPS-disciplined counter brings it into a
student lab. The provenance chain is what justifies the seventh
digit on the page.

\paragraph{Entry 5: mass loss in a long-running evaporation
experiment.}

\begin{description}[leftmargin=2em, itemsep=2pt]
  \item[Date / time.] 2026-04-01 09{:}00 through 2026-04-29 17{:}00,
    twice-daily mass readings (09{:}00 and 17{:}00) plus weekend
    single 12{:}00 readings.
  \item[Instrument.] Mettler-Toledo XPR205 analytical balance,
    capacity 220 \si{\gram}, readability
    $0.01 \,\si{\milli\gram}$, repeatability $0.015
    \,\si{\milli\gram}$ at full load
    \cite{web:ch01t3-mettler-xpr205}, serial no.\ B934517201,
    last cal.\ 2025-12-10 (in-house cal against OIML R 111 class E2
    weight set, NIST-traceable certificate MT-XPR205-251210; balance
    type-approval OIML R 76 class I). Internal motorised
    calibration weight engaged before each session.
  \item[Measurand and setup.] Mass loss of a $20.00 \,\si{\milli
    \liter}$ open beaker of reagent-grade ethanol left to evaporate
    in a fume hood at controlled draught.
  \item[Operator.] N.~P.~(chemistry undergraduate, summer project).
  \item[Conditions.] $T = 22.0 \pm 0.3 \,\degC$, RH
    $38 \pm 4 \,\si{\percent}$, fume hood face velocity
    $0.40 \,\si{\meter\per\second}$ (Sentry 800 anemometer, monthly
    cal.). Balance compartment doors closed for 60 s before each
    reading; vibration isolation table; balance compartment
    temperature logged at each reading via a Madgetech
    HiTemp140-PT placed beside the beaker.
  \item[Provenance of reference.] Mass scale traceable to the SI
    through the OIML R 111 E2 weight set; the kilogram itself is
    realised from the fixed Planck constant
    \cite{web:bipm-mise-en-pratique-kilogram-2019} through Kibble-balance
    primary realisations at NIST, NPL, and PTB.
  \item[Estimate before reading.] Vapour-pressure-driven evaporation
    rate for ethanol in still-ish air at $22 \,\degC$ gives
    $\sim 0.6 \,\si{\milli\gram\per\second}$, equivalent to
    $\sim 100 \,\si{\milli\gram}$ in 3 minutes and on the order of
    $50 \,\si{\gram}$ in 24 h. Recorded.
  \item[Reading.] Initial sample $15.794 \,\si{\gram}$; logged twice
    daily for 28 days; total mass loss
    $14.22 \,\si{\gram}$ over the period; instantaneous loss rate
    fitted as $-0.55 \,\si{\milli\gram\per\second}$ at day 1,
    falling to $-0.10 \,\si{\milli\gram\per\second}$ at day 28 as
    the residual fluid line dropped below the beaker rim. A clean
    deviation appeared at day 14 when the fume hood draught was
    inadvertently raised by a colleague (the resulting acceleration
    is logged and annotated; the hood face-velocity reading at the
    next session caught it).
  \item[Signature.] N.~P., 2026-04-29; PI signature R.~T.,
    2026-04-30.
\end{description}

The four-week duration is what reveals the day-14 hood-draught
disturbance; a single short measurement would have read it as drift
in the balance, in the reagent, or in the operator. The combination
of a stable balance and a co-located condition logger lets the
engineer correctly attribute the disturbance.

\paragraph{What the worked entries show.}

The five entries differ in instrument, reagent, hazard, and time
scale. They are identical in form. Every entry carries: a dated
reading, a named instrument with a serial number and a current
calibration certificate, a named operator, the conditions, a chain
of provenance back to a national standard, the reading itself, the
estimate that preceded it, and the operator's signature. A
notebook in which most entries omit two or three fields is not a
notebook in this sense; it is a diary.

\subsection{Reading the meter properly}

A practical aside, applicable to most analog meters and many
digital ones:

\begin{itemize}[leftmargin=*]
  \item Allow the instrument to stabilise before reading. Most
    instruments have a settling time the manual specifies; many have
    a longer time the manual does not.
  \item Read the meter with the eye perpendicular to the scale.
    Parallax error is the most easily missed source of bias.
  \item Note the zero offset before each reading session, then
    correct the reading or record that no correction was applied.
  \item For digital instruments, note the last-digit resolution and
    the manufacturer's stated noise floor; do not report a precision
    finer than either.
\end{itemize}

These practices are small enough to skip and important enough to
record here. They are the part of the discipline that a textbook can
name but only practice can make real.

\subsection{Historical sketch: the metric system, 1799--2019}\pathtag{standard}

The measurement habit named above relies on a chain of references
that the working engineer does not maintain personally. The chain
exists because two centuries of scientific and political work built
the institutions that hold it. A short history grounds why the habit
is more than personal hygiene and why the seventh field of the
schema, the provenance line, can be populated at all.

\paragraph{1799: the m\`etre des Archives.}

On 4 messidor an VII of the French Republican calendar (22 June
1799), the Commission temporaire des Poids et Mesures deposited a
platinum bar and a platinum cylinder at the Archives de la
R\'epublique in Paris \cite{hist:ch01t3-archives-metre-1799}. The bar
was 1 m\`etre long, defined as one ten-millionth of the quadrant of
the Earth's meridian from the equator to the pole passing through
Paris; the cylinder weighed 1 kilogramme, defined as the mass of one
cubic decim\`etre of pure water at the temperature of its maximum
density. The artefacts are today held at the Conservatoire National
des Arts et M\'etiers in Paris. They are the first physical
realisations of base units intended for international use rather
than for a single trade, city, or guild.

\paragraph{1875: the Metre Convention.}

On 20 May 1875, the Conf\'erence diplomatique du M\`etre concluded a
treaty signed by seventeen states establishing the Bureau
International des Poids et Mesures (BIPM) at S\`evres, just outside
Paris, together with the Comit\'e International des Poids et Mesures
(CIPM) and the Conf\'erence G\'en\'erale des Poids et Mesures (CGPM)
\cite{hist:ch01t3-metre-convention-1875}. The Convention created the
institutional spine the chapter has been silently leaning on. Every
modern calibration certificate that reads ``traceable to the SI''
runs back, through one or more national metrology institutes, to a
chain anchored at S\`evres.

\paragraph{1889: the IPK at S\`evres.}

The first General Conference on Weights and Measures, in 1889, took
delivery of thirty platinum-iridium cylinders (90\% Pt, 10\% Ir),
manufactured by Johnson Matthey of London and adjusted by the BIPM
\cite{hist:ch01t3-cgpm1-1889}. One was sanctioned as the
International Prototype of the Kilogram (the IPK, prototype K);
several others were assigned as national prototypes (prototype K20
for the United States, K38 for Germany, K18 for the United Kingdom).
The metre underwent a similar treatment: the International Prototype
of the Metre (IPM), a platinum-iridium X-section bar with three
fine scribed lines near each end, replaced the m\`etre des Archives.
The IPK kept the kilogram on an artefact basis for the next 130
years; periodic verifications at S\`evres recorded a slow drift of
the prototypes against their copies, on the order of a few tens of
micrograms per century, with no way to determine which had moved
\cite{paper:girard-ipk-1994}.

\paragraph{1948 and 1960: SI and the krypton metre.}

The 9th CGPM (1948) adopted the mean solar second (1/86400 of the
mean solar day) as the temporary definition of the unit of time and
asked CIPM to work out a coherent practical system suitable for
general adoption \cite{hist:ch01t3-cgpm9-1948}. The 11th CGPM (1960)
adopted that recommendation, named the practical system the Syst\`eme
International d'Unit\'es (SI), and redefined the metre as
$1\,650\,763.73$ wavelengths in vacuum of the orange-red line from
the $2p_{10} \to 5d_5$ transition of krypton-86
\cite{hist:ch01t3-cgpm11-1960}. The metre had become reproducible in
any sufficiently equipped laboratory; the prototype metre was now a
museum object.

\paragraph{1967: the caesium second.}

The 13th CGPM (1967) redefined the second as the duration of
$9\,192\,631\,770$ periods of the radiation corresponding to the
transition between the two hyperfine levels of the ground state of
the caesium-133 atom \cite{hist:ch01t3-cgpm13-1967}. With this
definition the second became, for the first time, a unit defined by
a property of matter accessible in any well-instrumented laboratory.
Caesium-fountain primary clocks built since the 1990s realise the
definition to better than $1 \times 10^{-15}$ in fractional
frequency; the worked notebook entry on mains frequency above
inherits its provenance from this chain.

\paragraph{1983: the speed of light made exact.}

The 17th CGPM (1983) redefined the metre as the length of the path
travelled by light in vacuum during a time interval of $1/299\,792\,
458$ of a second, fixing $c = 299\,792\,458 \,\si{\meter\per\second}$
exact by definition \cite{hist:ch01t3-cgpm17-1983}. From 1983 the
metre is realised whenever a frequency-stabilised laser tied to the
caesium second is available; gauge blocks and interferometric length
metrology run back to this definition.

\paragraph{2018: the kilogram released from the artefact.}

The 26th CGPM, meeting in Versailles 13--16 November 2018, adopted
Resolution 1 redefining the kilogram, the ampere, the kelvin, and the
mole through fixed values of four defining constants: the Planck
constant $h = 6.626\,070\,15 \times 10^{-34} \,\si{\joule\second}$,
the elementary charge $e = 1.602\,176\,634 \times 10^{-19}
\,\si{\coulomb}$, the Boltzmann constant $k = 1.380\,649 \times
10^{-23} \,\si{\joule\per\kelvin}$, and the Avogadro constant
$N_A = 6.022\,140\,76 \times 10^{23} \,\si{\per\mole}$, with effect
from 20 May 2019 \cite{hist:ch01t3-cgpm26-2018}. After 2019, every
base unit of the SI is defined through a fixed constant of nature;
the IPK, after 130 years as the world standard for mass, became a
metrology artefact of historical rather than functional importance.

\paragraph{Why the habit needed the standards to lean on.}

The notebook entry on the evaporation experiment cites a chain that
runs from a Mettler-Toledo balance through an OIML R 111 E2 weight
set to the Kibble-balance realisations of the kilogram at NIST, NPL,
and PTB, which in turn realise the unit through a fixed Planck
constant per the 2018 resolution. The chemistry titration entry
cites NIST SRM 84n, which in turn ties to the mole through the
Avogadro constant. The frequency entry cites a GPS-disciplined OCXO
locked to the USNO master clock, traced through caesium-fountain
primaries to the second defined by the 1967 caesium-133 transition.
None of these chains were available to the engineer of 1799. The
habit named in this section is what it is because the institutions
above exist and because the constants of nature on which the modern
SI rests are reproducible in any well-equipped national metrology
institute. The habit is not personal hygiene; it is what every
working engineer does to make the institutional chain reach the
bench. The full traceability picture at one glance is reproduced as
Figure~\ref{fig:vol01:ch01:traceability-pyramid}, with forward
detail in Chapter 3.

\input{volumes/01-quantity/ch01-why-we-measure/figures/fig-traceability-pyramid.tex}

\section{Failure: the metric vs.\ imperial fault line}\pathtag{core}

Every chapter in this work has a failure section. We begin with a
chronic failure mode: unit systems meet at interfaces, and the
conversion is treated as memory work.

SI is the international system maintained through the Bureau
International des Poids et Mesures and the General Conference on
Weights and Measures \cite{std:bipm-si2019}. US customary and
imperial units persist in specific jurisdictions, codes, installed
instruments, and legacy designs. The persistence has engineering
causes: a retrofit of every imperial datum to SI would invalidate a
continent of drawings, calibrations, parts catalogues, machine tools,
and certified procedures. The transition has happened in industries
where the value of consistency exceeded the cost of conversion
(international science, modern aerospace design, pharmaceuticals);
it has not happened where the cost is high and the conservatism is
institutional. For the foreseeable future, the engineer will live in
a two-system world.

The two-system world has a chronic failure mode. We saw it acutely in
section 1.4 of this chapter; it is also present, every day, in
quiet ways:

\begin{itemize}[leftmargin=*]
  \item A drawing in inches is interpreted in millimetres.
  \item A torque specification in newton metres is read as
    foot pounds.
  \item A fuel quantity in pounds per litre is treated as kilograms
    per litre.
  \item A pressure rating in psi is converted to bar with the wrong
    coefficient.
  \item A medical dose in micrograms is read as milligrams.
\end{itemize}

These errors look clerical until they reach a patient, aircraft,
plant, or bridge. Preventing them requires written units, checked
conversions, and audited interfaces. We state it as a rule we use
throughout this work.

\begin{principle}[Units carry through every step]
Every quantity in every calculation, every entry in every log, every
field in every interface carries its unit explicitly. Conversion
between systems happens at named, audited boundaries. Within a
calculation, the dimensions cancel; if they do not, the calculation
is wrong before any number is computed.
\end{principle}

\input{volumes/01-quantity/ch01-why-we-measure/figures/fig-unit-interface.tex}

The book is written in SI throughout. Where a real artifact uses
imperial units (a US-built reactor pressure vessel, a Boeing 737
fastener size, a UK plumbing fitting, an inch-based machine tool),
we give the imperial value parenthetically as the artifact's
working number, and the SI value as the engineering quantity.
Conversions are done in writing, not in the reader's head.

We close on this rule because it is the measurement habit the reader
will use in every later chapter. The reader who carries it forward
has already done a serious part of an engineer's work.

\section{Measurement as the bond of accountability}\pathtag{standard}

A measurement is not only a description of the world; it is an
artifact in a chain of accountability. Three further fault lines
recur in working engineering, each rooted in measurement and each
visible only when a chapter takes its time over them: a
sampling-protocol failure that recalled 36 million pounds of food,
the compliance frameworks under which a measurement either counts
or does not count, and a working audit checklist that a reader can
apply to any measurement entered into a log.

\subsection{Applied vignette: a measurement read back seven years later}\pathtag{standard}

A measurement's value as a bond of accountability is easiest to see
when the bond is called. Consider a concrete pour whose record is
opened not on the day it was cast but seven years later, when a hairline
crack in a car-park deck slab prompts a structural review.

The reviewing engineer pulls the batch record for grid line C, pour
sequence 14, dated 2019. The record carries what the measurement habit
requires: a delivery ticket naming mix design C40/50, a slump of
$110 \,\si{\milli\meter}$ measured with a cone whose calibration date
is on the ticket, a chloride content below the specified limit, and
four cylinders cast on site. The 28-day cylinder results are logged at
$52.4$, $53.1$, $51.8$, and $48.9 \,\si{\mega\pascal}$, each with the
cylinder identity, the testing machine's serial number and calibration
certificate, the technician's initials, and the curing-tank
temperature. Three cylinders sit comfortably above the specified
$50 \,\si{\mega\pascal}$; the fourth, at $48.9 \,\si{\mega\pascal}$, is
flagged in the record with a note that the cylinder showed a capping
defect and was reported for information rather than acceptance, per the
governing test standard's outlier provision.

The record answers the question the crack raised. The concrete met its
strength specification; the low cylinder was a capping artefact, not a
weak pour; the chloride content rules out the reinforcement corrosion
that would otherwise be the first suspect for a deck slab. The review
turns to loading history and thermal movement instead, and closes the
concrete as a cause within a day.

Now consider the same slab with a different record: a delivery ticket
with the mix design but no slump, cylinder results written as
``$\approx 50$, all passed'' with no cylinder identities, no testing
machine named, no technician, no note on the low result because the low
result was never separately recorded. The concrete may have been
identical. The record cannot say so. The reviewing engineer cannot
exclude a weak pour, cannot trace the testing machine's calibration,
cannot distinguish a capping defect from a genuine low strength, and
cannot close the concrete as a cause. The investigation widens, cores
are drilled, and the cost of the missing fields is paid seven years
after the thirty seconds it would have taken to record them. The two
records describe the same slab; only one of them is engineering
evidence.

\subsection{Measurement-protocol failure: the Cargill ground-turkey recall, 2011}\pathtag{standard}

\paragraph{The case.} Between 20 February and 2 August 2011, the
Cargill Meat Solutions plant at Springdale, Arkansas, produced
approximately $35{,}900{,}000$ pounds ($16{,}300{,}000 \,\si{\kilogram}$,
roughly 36 million pounds in round figures) of fresh and frozen
ground turkey. A fraction of that production was intermittently
contaminated with a multidrug-resistant strain of \emph{Salmonella
enterica} serotype Heidelberg. The strain was characterised by a
pulsed-field gel electrophoresis (PFGE) pattern designated
JF6X01.0223 in the Centers for Disease Control and Prevention's
PulseNet database, and showed resistance to ampicillin,
streptomycin, tetracycline, and gentamicin, with reduced
susceptibility to ceftriaxone, complicating clinical management of
severe cases \cite[p.~46]{acc:ch01t3b-cdc-mmwr-salmonella-heidelberg-2013}.

The CDC linked the strain to ground turkey purchased by
case-patients across 34 US states through PulseNet pattern
matching. The confirmed-case count at the close of the outbreak
investigation was 136 illnesses, with at least 37 hospitalisations
and one death (Carmen Garibay-Gonzalez, of California). Three
positive retail-sample isolates and one positive in-plant isolate
were traced back to the Springdale plant, three of which carried
the outbreak PFGE pattern. On 3 August 2011, the U.S. Department of
Agriculture's Food Safety and Inspection Service (FSIS) issued
recall release 090-2011, requiring Cargill to recall the entire
production of the plant for the period \cite{acc:ch01t3b-usda-fsis-recall-090-2011}.
After Cargill resumed operation at the same plant later that month,
a second cluster appeared; FSIS issued recall expansion 094-2011 on
11 September 2011, covering an additional $185{,}000$ pounds of
ground turkey produced on 23, 24, 30, and 31 August 2011 \cite{acc:ch01t3b-usda-fsis-recall-094-2011}.

The combined recall mass of approximately 36 million pounds was, at
the time, the largest meat recall in US history measured in product
mass. Cargill reported the direct recall cost at approximately
$\$30 \,\text{million}$ in regulatory and industry disclosures;
downstream costs (lost production at the idled plant, brand
recovery, litigation reserves) raised the total to a higher figure
that the company did not separately publish.

\paragraph{The measurement story.} The reason the contamination
escaped routine federal sampling for months was not analytical. The
laboratory work was correct. Every individual sample that FSIS
collected was a faithful reading of the volume of product from
which it was drawn. The failure was at the level of the protocol:
the granularity at which samples were aggregated did not match the
granularity at which the product could be contaminated.

FSIS's pre-2011 ground-turkey sampling under the Pathogen Reduction
and Hazard Analysis and Critical Control Points (PR/HACCP)
framework collected composite samples at the establishment level,
on a frequency tied to the plant's production volume and the
inspector's schedule. Each sample was treated as representative of
the entire production day at the plant, not of any individual
consumer-package lot \cite[pp.~12--18]{acc:ch01t3b-gao-fsis-sampling-2012}.
A plant of Springdale's output rate, approximately $400{,}000$ pounds
per day, produced thousands of consumer-package lots per production
day. A single composite sample drawn from one production-day stream
could clear the day's output regardless of whether contamination
was confined to a particular grinder line, a particular shift, or a
particular vendor of incoming birds.

We can put a number to the granularity mismatch. If a Springdale
production day produced $\sim 400{,}000$ pounds of ground turkey
delivered to consumers in $\sim 1$-pound packages, the lot count
per day was $\sim 4 \times 10^{5}$ packages. A single composite
sample collected from a representative point in the production
stream covers, in the limit, a single point. Even with several
samples per shift and an assumption that contamination spreads
homogeneously across the day's product within a plant, the implicit
coverage ratio of inspected mass to consumed mass for any
within-day failure mode is order $10^{-6}$. The protocol assumed
homogeneity at a granularity (the production day) at which the
process is, in practice, not homogeneous.

A retrospective review by the U.S. Government Accountability Office
found that the outbreak strain had been isolated from Cargill
ground turkey samples collected by USDA-FSIS, the National
Antimicrobial Resistance Monitoring System (NARMS), and retail
testing at multiple points during 2011 before the recall, but no
single sample triggered a recall because each was treated as a
stand-alone establishment-level finding rather than as one entry in
a longitudinal record tied to lot-level traceback \cite[p.~22]{acc:ch01t3b-gao-fsis-sampling-2012}.
The samples existed, and were correct; the linkage between them
that would have triggered earlier recall did not.

The case became canonical in food-engineering and process-safety
literature for a principle that recurs across domains:

\begin{principle}[Sampling granularity matches failure granularity]
A measurement protocol is correct only if the granularity at which
samples are aggregated matches the granularity at which the product
can fail. Sampling at the plant-day level cannot catch contamination
confined to a shift, a line, a vendor, or a lot, no matter how many
samples are taken or how accurately each is analysed.
\end{principle}

Following the recall, FSIS issued updated guidance moving toward
more granular sampling streams, tighter linkage between PulseNet
pattern matches and recall triggers, and a 2013 framework requiring
earlier coordination between FSIS, the CDC, and state health
departments when PulseNet detected an antibiotic-resistant
\emph{Salmonella} cluster. The pattern is the inverse of the Mars
Climate Orbiter failure: there, every individual measurement was
honest but the units changed at an unaudited boundary; here, every
individual measurement was honest but the boundary at which samples
were aggregated did not match the boundary at which failure could
occur. Both are protocol failures masquerading as measurement
successes.

\subsection{Measurement in compliance: cGMP, ISO 9001, 14 CFR 25.1309}\pathtag{standard}

A measurement that gives the correct value, in the correct units,
with the correct uncertainty, can still fail compliance review.
Engineering compliance is the discipline by which a measurement is
not only correct but also \emph{counted}: documented in the form
the regulator or accreditor will accept, traced to the framework
under which the artifact is approved, and signed by an authority
the framework recognises. We name three frameworks the reader will
encounter throughout this book.

\paragraph{cGMP for pharmaceuticals.} In the United States, the
Food and Drug Administration's current good manufacturing practice
(cGMP) regulations for finished pharmaceuticals are codified in
Title 21 of the Code of Federal Regulations, Parts 210 and 211
\cite{law:ch01t3b-21cfr210-211}. Part 211 Subpart I, ``Laboratory
Controls'' (\S\,211.160 through \S\,211.176), is the
measurement-and-records backbone of pharmaceutical manufacturing
compliance. Subpart I requires that laboratory controls
``shall include the establishment of scientifically sound and
appropriate specifications, standards, sampling plans, and test
procedures'' (\S\,211.160(b)); that the calibration of laboratory
instruments be performed against ``a known standard with a
traceability to a national or international standard, when
available'' (\S\,211.160(b)(4) via FDA guidance); that all data
secured in the course of laboratory tests be ``recorded at the
time they are performed'' and that ``[a]ny deviation from the
written specifications shall be recorded and justified''
(\S\,211.194(a)). An FDA Form 483 (the inspector's observation
form) is most often written for failures of this clause: a correct
analytical reading that was not contemporaneously recorded, or that
was recorded but not signed, or that was signed but not by the
analyst who performed it. The measurement gives the right value;
the record does not stand up under audit.

\paragraph{ISO 9001 for general manufacturing.} ISO 9001:2015,
\S\,7.1.5 ``Monitoring and measuring resources,'' is the
international analogue of the cGMP laboratory-controls clause for
general quality-managed manufacturing \cite{std:ch01t3b-iso9001-2015}.
The clause requires that the organisation:

\begin{itemize}[leftmargin=*]
  \item determine the measurement resources needed (instruments,
    standards, software);
  \item ensure those resources are ``suitable for the specific
    type of monitoring and measurement activities being
    undertaken'';
  \item maintain them, with records, ``to ensure their continued
    fitness for purpose'';
  \item where measurement traceability is a requirement (or
    deemed necessary by the organisation), verify and calibrate
    the measuring equipment at specified intervals or before use,
    against measurement standards traceable to international or
    national measurement standards;
  \item where no such standards exist, retain ``the basis used for
    calibration or verification'' as documented information;
  \item identify equipment status, safeguard it from adjustments
    that invalidate the calibration, and take action when the
    equipment is found to be unfit for its purpose, ``including
    determining and assessing the validity of previous measurement
    results.''
\end{itemize}

The last clause is the audit teeth. A calibration found out of
tolerance triggers an obligation to evaluate every measurement
taken with that instrument since its last successful calibration.
A manufacturer that does not maintain a record of which
measurements were taken with which instrument cannot satisfy the
clause; an ISO 9001 surveillance audit will write a major
non-conformity.

\paragraph{14 CFR 25.1309 for transport aircraft.} The U.S.
Federal Aviation Administration's certification rule for
transport-category airplanes, Title 14 CFR \S\,25.1309, is the
reliability-by-measurement framework for civil aviation
\cite{law:ch01t3b-14cfr25-1309}. The rule requires that each system
be ``designed and installed so that\dots the occurrence of any
failure condition which would prevent the continued safe flight
and landing of the airplane is extremely improbable, and\dots the
occurrence of any other failure condition which would reduce the
capability of the airplane or the ability of the crew to cope with
adverse operating conditions is improbable.''

The interpretive guidance, Advisory Circular AC 25.1309-1A, sets
quantitative probability targets:

\begin{itemize}[leftmargin=*]
  \item \emph{Catastrophic} failure conditions (those resulting in
    hull loss with multiple fatalities) must be ``extremely
    improbable,'' interpreted as a probability of occurrence no
    greater than $\sim 10^{-9}$ per flight hour.
  \item \emph{Hazardous} failure conditions (those producing a
    large reduction in safety margins, serious or fatal injuries
    to a small number of occupants) must be ``extremely remote,''
    interpreted as $\sim 10^{-7}$ per flight hour.
  \item \emph{Major} failure conditions (significant reduction in
    safety margins, discomfort to occupants, possible injuries)
    must be ``remote,'' interpreted as $\sim 10^{-5}$ per flight
    hour.
\end{itemize}

The rule is enforced by a measurement chain. Each component's
mean-time-between-failures is measured in qualification testing,
service history, or analytical reliability modelling; the
combination of components into a system is propagated through fault
tree analysis or failure mode and effects analysis (FMEA); the
result is a system-level probability that must meet the rule. A
manufacturer that designs to the catastrophic target without a
documented measurement chain back to component-level reliability
data cannot certify the system. The Boeing 737 MAX MCAS
certification record is the canonical recent failure of this chain
at a single point: the system-level hazard analysis classified MCAS
runaway as ``major'' (target $10^{-5}$) when, as flown, it was in
practice a ``catastrophic'' single-source failure mode (target
$10^{-9}$).

\paragraph{The compliance principle.} A measurement that meets the
specification but lacks the documentation required by the
applicable compliance framework fails compliance review. The
analytical result is not in dispute; the recordkeeping is. The
working engineer learns, often the hard way, that a measurement
exists for the regulator and the auditor as much as it exists for
the physics. Both readers must be served by the same notebook
entry.

\subsection{The 7-question working audit checklist}\pathtag{standard}

A measurement entered into a working log can be audited along
seven independent axes. The seven questions below are the
operational reading of ISO/IEC 17025:2017, \S\,7.5 (``Technical
records'') and \S\,7.8 (``Reporting of results'')
\cite{std:ch01t3b-iso17025-2017}, distilled into a form a working
laboratory uses on every entry. Any entry that fails any one of the
seven is at risk in a compliance audit.

\begin{enumerate}[leftmargin=*]
  \item \textbf{What was measured?} The quantity, in SI units (or
    in the recorded units of the applicable code), with its
    value, its uncertainty, and its time stamp to the resolution
    the measurement supports. ``$237.4 \pm 0.1 \,\si{\milli\meter}$
    at 14:32:17 on 7 March 2025'' satisfies the question;
    ``$237.4 \,\si{\milli\meter}$'' alone does not, because no
    uncertainty and no time are recorded.

  \item \textbf{What instrument?} The make, model, and unique
    serial number or asset tag of the instrument used. A
    laboratory with three calibers of the same model that does
    not record which caliper took the reading cannot trace a
    later non-conformance to a specific instrument.

  \item \textbf{What calibration certificate?} The certificate
    number, the date of issue, the date of expiry, the
    accrediting body (NIST, NPL, PTB, INMETRO, A2LA, UKAS, or
    equivalent), the standard the instrument was traced to, and
    the uncertainty of the calibration. ISO/IEC 17025:2017
    \S\,7.8.5 requires that all of these appear on the
    certificate; the laboratory's audit log must reference it.

  \item \textbf{What operator?} The signature, initials, or
    user-ID of the person who took the reading. The standard
    requires the recorded identity be unambiguously linkable to a
    single person; ``J. Smith'' in a laboratory with three
    J. Smiths is insufficient.

  \item \textbf{What conditions?} Temperature, humidity,
    barometric pressure, vibration level, electromagnetic
    environment, and any other ambient parameter outside the
    instrument's calibration envelope. For dimensional
    measurement on a coordinate-measuring machine, the standard
    measurement temperature of $20 \,\degC \pm 1 \,\si{\kelvin}$
    must be confirmed at the time of measurement; deviation
    requires a recorded thermal correction.

  \item \textbf{What method?} The standard test method or
    standard operating procedure (SOP) under which the
    measurement was performed, by document number and revision:
    ``ASTM E384-22, Section 8.4''; ``SOP-QC-014 Rev.\ 7''. A
    measurement made under no recorded method cannot be compared
    to a measurement made elsewhere.

  \item \textbf{What signature?} A countersignature by a second
    qualified analyst or supervisor for critical measurements
    (release tests, in-process critical control points, change
    control). The signature must be attributable, dated, and
    accompanied by the role of the signer.
\end{enumerate}

\paragraph{Failure modes.} In a working audit each question fails
in characteristic ways:

\begin{itemize}[leftmargin=*]
  \item \emph{Missing certificate.} The instrument was in use,
    the readings are on file, but no calibration certificate is
    on hand for the relevant period. The auditor cannot
    distinguish a faithful reading from instrument drift. All
    readings taken with the instrument since the last documented
    calibration are quarantined pending resolution.

  \item \emph{Expired certificate.} The certificate is on file
    but lapsed before the reading was taken. ISO 9001:2015
    \S\,7.1.5.2 requires evaluation of measurements taken with
    out-of-calibration equipment; the laboratory must document
    the evaluation or quarantine the affected results.

  \item \emph{Mis-traced instrument.} The certificate cites a
    reference standard that is not itself traceable to the SI
    through an unbroken chain. Discovered most often in
    secondary or tertiary laboratories that have outsourced
    calibration to a provider lacking ILAC mutual-recognition
    accreditation.

  \item \emph{Environmental condition out of cal range.} The
    instrument's calibration certificate states the conditions
    under which it is valid (a thermal envelope, a humidity
    range). Readings taken outside that envelope require an
    additional correction or a re-calibration; readings taken
    outside the envelope without correction are unreliable.

  \item \emph{Operator unfamiliar with method.} The signed
    operator does not appear on the laboratory's training
    matrix as qualified to perform the cited method. ISO/IEC
    17025:2017 \S\,6.2.6 requires documented competence; the
    auditor will quarantine results.

  \item \emph{Signature not attributable.} A signature appears
    but cannot be linked to a single person with a single role
    on a single date. Common with shared user-IDs in legacy
    laboratory information management systems; the modern remedy
    is per-user authentication with audit-trail logging.
\end{itemize}

\paragraph{Audit outcomes.} A laboratory audit applies the
checklist to a random sample of entries and reports each entry as
\emph{pass}, \emph{fail}, or \emph{conditional}. A pass entry has
all seven questions answered. A fail entry has one or more
unrecoverable gaps (missing certificate, missing signature,
unidentifiable operator) and is grounds for a finding. A
conditional entry has a recoverable gap (a calibration certificate
that is on file but not linked to the entry, an out-of-envelope
condition with a recoverable correction available) and is
remediated within an agreed period, typically 30 or 90 days. In a
calibration laboratory accredited to ISO/IEC 17025:2017, the
non-conformance rate against the seven-question audit is the
single most-watched internal quality metric; a sustained rate above
$\sim 2\%$ on randomly sampled entries triggers a corrective-action
review with the accrediting body. The cost of remediation is
typically much higher than the cost of asking the seven questions
at the time the measurement is taken; this is the working
laboratory's daily reason for the discipline.

\begin{principle}[A measurement is an artifact in a chain of accountability]
A measurement is correct only if it satisfies the seven-question
audit: what, what instrument, what certificate, what operator,
what conditions, what method, what signature. The analytical value
is one element of a record; the record exists for the regulator
and the auditor as much as for the physics. A measurement that
fails any of the seven is not a measurement; it is a number on
paper.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Instrument the home, one week]
The reader instruments their immediate environment with the tools
they have, keeps a measurement notebook for seven consecutive days,
and produces a 1500-word reflection on what they learned.

\textbf{Track}. Household, with quantitative analysis. No physical
build; no chemical work; no electrical work above a battery.
\textbf{Hazard class}: none.
\textbf{Tools}. A kitchen scale (resolution $\sim 1 \,\si{\gram}$ for
small masses, $\sim 5 \,\si{\gram}$ for large). A thermometer
(kitchen, body, or outdoor). A stopwatch (any phone). A ruler or
measuring tape. A smartphone, used for: stopwatch, light-meter app,
sound-meter app (where available), accelerometer logging app, and
camera.

\textbf{Procedure}. Each day, choose at least three quantities to
measure. Suggested first-day list: ambient temperature in three
rooms; mass of a meal before and after cooking; the exact time the
kitchen tap takes to fill a $1 \,\si{\liter}$ container at full
flow; the pulse rate at rest and after one flight of stairs; the
horizontal distance walked in 30 paces. Continue with quantities of
the reader's choice for six more days.

\textbf{Record}. Each entry must contain: date, time, instrument
(name and any visible serial number), operator (the reader's
initials), conditions, the reading, and an estimate the reader made
\emph{before} taking the reading. Where the reader could not
estimate, that is itself recorded.

\textbf{Analysis}. For at least three of the quantities, the reader
takes ten or more measurements and computes (i) the mean, (ii) the
sample standard deviation, (iii) the implied $\pm 1\sigma$ uncertainty
on a single reading, (iv) the implied uncertainty on the mean. For at
least one quantity, the reader compares two different instruments
(say, the kitchen thermometer and the smartphone weather app) and
records the disagreement.

\textbf{Reflection (1500 words)}. The reader writes a 1500-word
narrative addressing:

\begin{itemize}[leftmargin=*]
  \item Which quantities were stable and easy to measure, and which
    drifted.
  \item Where the reader's estimate was within a factor of three of
    the measurement, and where it was off by more.
  \item Which quantities the reader could not measure with the tools
    available, and what would be needed to measure them.
  \item Where two instruments disagreed and how the reader judged
    which to trust.
  \item One quantity in the reader's home that an engineer would
    have wanted to measure but the reader did not.
\end{itemize}

\textbf{Deliverable}. The notebook (photographed or scanned) plus
the reflection. The general editor's reference logbook for this
project is published online; the reader is encouraged to compare
\emph{after} completing their own. There is no single correct
answer.

\textbf{Time}. Twenty to thirty minutes per day for seven days, one
repeat-measurement session of one to two hours, plus two to three
hours of analysis and writing.

\textbf{Phases}.
\begin{enumerate}[leftmargin=*]
  \item \emph{Day 1: orientation}. Estimate-then-measure for the
    suggested first-day list. The point is to calibrate the
    reader's intuition against the first ground truth.
  \item \emph{Days 2--5: free choice}. Three to five quantities
    per day, including at least one outside the kitchen. By Day 3
    the reader should pick at least one quantity that they expect
    to be hard to measure with the tools at hand.
  \item \emph{Day 6: repeat-measurement campaign}. Ten or more
    measurements of the same quantity in succession, then again
    six to eight hours later. Computes the sample standard
    deviation and the implied $\pm 1\sigma$ uncertainty.
  \item \emph{Day 7: instrument cross-check}. Two instruments
    against the same quantity. Document agreement and
    disagreement.
\end{enumerate}

\textbf{Success criteria}.
\begin{enumerate}[leftmargin=*]
  \item Notebook is complete: every entry carries date, time,
    instrument identity, operator initials, conditions, reading,
    and the prior estimate.
  \item For at least three quantities, the reader has computed
    mean, standard deviation, single-reading $\pm 1\sigma$, and
    standard uncertainty of the mean, with correct arithmetic
    (verified against the formulas in Chapter 4 once it is read).
  \item At least one cross-check between two instruments is
    documented with a numerical disagreement and a written
    judgement about which to trust and why.
  \item Reflection addresses all five prompts. A reflection that
    leaves one bullet unaddressed is incomplete.
  \item The reader can name at least one quantity in their home
    that they could not measure and what instrument would change
    that.
\end{enumerate}

\textbf{Common failure modes}.
\begin{itemize}[leftmargin=*]
  \item Skipping the prior-estimate column on busy days. The
    estimate column is the experiment; without it the project
    cannot calibrate the reader's intuition.
  \item Choosing only easy quantities. If every entry is a
    kitchen mass or a stopwatch interval, the reader will not
    encounter the discomfort the chapter is teaching.
  \item Treating disagreement between two instruments as
    instrument failure rather than data. Disagreement is the
    measurement, not its absence.
  \item Recording instrument names without serial numbers or
    model numbers. ``A kitchen thermometer'' is a class; the
    notebook needs the instance.
  \item Writing readings to more digits than the instrument
    supports. A $\$10$ thermometer that reads in $1 \,\degC$
    increments cannot, by writing $22.4 \,\degC$, suddenly read
    to $0.1 \,\degC$.
  \item Conflating mean and reading. Ten samples of a tap-fill
    time give the engineer four numbers (mean, standard
    deviation, $\pm 1\sigma$ on a single reading, $\pm 1\sigma$
    on the mean) where the notebook still records a single
    instantaneous reading per row.
  \item Omitting the conditions field when the conditions
    obviously did not change. Drift across a week of readings is
    invisible without a condition log.
\end{itemize}

\textbf{Scoring rubric for the deliverable}.
The project is assessed against four rubric items, each scored
from $1$ (poorly addressed) to $5$ (textbook-quality). A submission
scores a maximum of $20$; a passing submission scores at least
$12$ with no individual item below $2$.

\begin{enumerate}[leftmargin=*]
  \item \emph{Field completeness} ($1$--$5$). Every notebook entry
    populates the six fields of the schema: date / time, instrument
    (model + serial or unique identifier), operator, conditions,
    reading, prior estimate. Entries that omit one field
    consistently lose one point; entries that omit two or more lose
    two.
  \item \emph{Instrument traceability} ($1$--$5$). For at least
    half of the instruments used, the reader documents the
    calibration chain at least one step beyond the instrument:
    manufacturer datasheet, factory-cal certificate, last
    calibration date, or chain to a national lab. A submission
    that names only the instrument scores $1$--$2$; a submission
    that documents a tie to a national standard for at least one
    instrument scores $4$--$5$.
  \item \emph{Quantitative uncertainty} ($1$--$5$). For at least
    three of the quantities, the reader computes mean, sample
    standard deviation, $1\sigma$ single-reading uncertainty, and
    $1\sigma$ uncertainty on the mean, with the arithmetic shown
    or verifiable in code. A submission that reports only mean
    and range scores $2$; a submission with all four numbers
    correctly derived and shown scores $5$.
  \item \emph{Provenance signature} ($1$--$5$). Each notebook
    entry is signed (or initialed-and-dated) by the operator; a
    cross-check entry is countersigned by a second reader where
    available; the reflection notes any disagreements between
    instruments and how the reader judged. A submission with no
    signatures scores $1$; a submission where every entry carries
    a signature and at least one cross-check has been
    countersigned scores $5$.
\end{enumerate}

\textbf{Worked submission examples}.

\textit{High-scoring submission (score 18/20).}
Seven days of entries, one to two facing pages per day, on graph
paper, photographed at the end of the week. Each entry carries
date and time stamped at the top; each instrument is named with
model and serial number where visible (kitchen scale Salter 1066
SVDR8, oven thermometer Polder THM-378, smartphone iPhone 13 mini
serial \texttt{C8PXVL2DQ1JK} for stopwatch and light-meter app
``Galactica Lux'' v.\ 3.2.1); the reader's initials sit at the
bottom of every entry. The cross-check on Day 7 compares the
kitchen oven thermometer against the smartphone's IR-thermometer
attachment (a Seek Thermal CompactPRO), records a
$+6 \,\degC$ disagreement at $180 \,\degC$ oven setpoint,
investigates, and concludes that the smartphone IR is reading
the oven glass rather than the cavity (a calibration-condition
mismatch). The repeat-measurement campaign on Day 6 logs twelve
fill times for a $1 \,\si{\liter}$ container at the kitchen tap;
the reader computes $\bar t = 8.42 \,\si{\second}$, $s = 0.18
\,\si{\second}$, $\sigma_{\bar t} = 0.052 \,\si{\second}$, and
notes that the morning batch was systematically $0.3
\,\si{\second}$ slower than the evening batch. A 1500-word
reflection addresses every prompt, notes a calibration chain
back to NIST for the Polder thermometer (factory cal cert.\
photographed), and identifies indoor RH as the quantity the
reader could not measure without a hygrometer.

Rubric: completeness $5$, traceability $4$ (kitchen scale has no
documented chain), uncertainty $5$, signature $4$ (cross-check
not countersigned, but every entry initialed). Total $18/20$.

\textit{Low-scoring submission (score 7/20).}
Five days of entries on loose-leaf paper, one to three readings
per day. Date present; time only on three entries. Instruments
named by class only (``kitchen thermometer,'' ``smartphone
stopwatch''). No serial numbers; no calibration history; no
condition logging beyond ``in kitchen.'' Prior estimate populated
on Day 1 only. Repeat measurements absent. A 600-word reflection
addresses two of the five prompts and concludes that ``the
project went fine.'' No signatures.

Rubric: completeness $2$, traceability $1$, uncertainty $1$,
signature $1$. Total $5/20$, below the passing floor; the
submission would be returned with a request to add the missing
fields and the repeat-measurement campaign, not failed outright.

The rubric is not a grading instrument; it is a way for the
reader to read their own submission against the chapter's
intent before turning it in.
\end{project}

\section{Exercises}

The exercises mix calculation, derivation, estimation, simulation,
design, diagnosis, reverse engineering, failure analysis, and
judgment per the editorial settlement on exercise types (Q16). Full
solutions are provided for the calculation and derivation exercises;
the estimation exercises receive published reference estimates with
the editor's working; the design, diagnosis, failure-analysis, and
judgment exercises receive rubrics rather than single answers.
Exercises are tagged with their reader-path tier: \emph{core} are
required, \emph{standard} are working-engineer default, and
\emph{mastery} use tools (covariance, programming, web data) the
chapter has not yet taught and may be deferred to chapter 4 or 8.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Convert $1\,\text{ft}$ to metres exactly. State your answer to four
significant figures, then state the exact value (the foot is defined
as $0.3048 \,\si{\meter}$). Comment on whether ``four significant
figures'' is the right report.

\paragraph{Solution.}
We compute the value, name the source of its exactness, distinguish
defined from measured quantities, and state the engineering
convention for reporting.

\textit{Step 1, write the four-significant-figure value.}
$1 \,\text{ft} = 0.3048 \,\si{\meter}$ to four significant figures.
This is the same value as the exact definition; the rounding to four
figures has not lost any digits.

\textit{Step 2, identify the source of the exactness.} The
international foot is defined as exactly $0.3048 \,\si{\meter}$ by the
International Yard and Pound Agreement of 1959, ratified by the
United States, the United Kingdom, Australia, Canada, New Zealand,
and South Africa. The defined value is a constant of the unit system
\cite{std:nist-sp811}; it is exact in the same sense that
$1 \,\si{\meter} = 1 \,\si{\meter}$ is exact.

\textit{Step 3, distinguish defined from measured.} The international
foot is a defined conversion factor: the foot is defined in terms of
the metre, with no further measurement required. Compare with the
older Imperial foot (the standard before 1959), which was defined
through a physical artefact (the Imperial Standard Yard) and was
therefore subject to measurement uncertainty; its value with respect
to the metre was approximately $1 \,\text{ft} \approx 0.30479997
\,\si{\meter}$, with a residual uncertainty in the seventh decimal
place from the artefact intercomparison. The 1959 redefinition
eliminated the artefact uncertainty by fixing the foot by definition.

\textit{Step 4, evaluate the significant-figures language.} The
phrase ``four significant figures'' implies an uncertainty of order
$\pm 0.00005 \,\si{\meter}$. The defined value has zero uncertainty,
so the implication is wrong. The four-figure form is not less
precise than the exact value; it carries the same information at a
lower display resolution.

\textit{Step 5, identify the working convention.} The convention for
defined constants in engineering reports is to state the definition
and avoid the significant-figure framing. ``The foot is defined as
exactly $0.3048 \,\si{\meter}$,'' or ``$1 \,\text{ft} \equiv 0.3048
\,\si{\meter}$,'' communicates the exactness directly. The
significant-figures language is reserved for measured quantities,
where it carries genuine information about the measurement
uncertainty.

\textit{Step 6, name the related cases.} The same convention applies
to several other widely-used conversions that are defined-by-
international-agreement: $1 \,\text{lb} = 0.45359237 \,\si{\kilogram}$
exactly; the IT-calorie at $4.1868 \,\si{\joule}$ exactly; the
nautical mile at $1852 \,\si{\meter}$ exactly. Each is an
international-agreement constant; each is reported as exact rather
than as a many-digit approximation.
\end{exercise}

\begin{exercise}
The published mass of a Boeing 747-400 at maximum take-off weight is
approximately $396 \,\si{\tonne}$. Express this in (a) kilograms,
(b) pounds, (c) slugs. Identify which of those a US flight crew, a
US ground engineer, and a European fuelling crew would each
plausibly use.

\paragraph{Solution.}
We work each part as a multiplication by an exact ratio equal to one,
keeping the unit symbols on the right of every numerator.

\textit{Step 1, write the SI value.} The metric tonne is defined as
$10^{3} \,\si{\kilogram}$, so
$396 \,\si{\tonne} = 3.96 \times 10^{5} \,\si{\kilogram}$. This is an
exact conversion within SI; no rounding has occurred.

\textit{Step 2, convert to pounds.} The international avoirdupois pound
is defined as $0.45359237 \,\si{\kilogram}$ exactly
\cite{std:nist-sp811}, so
$1 \,\si{\kilogram} = 1/0.45359237 = 2.2046226\,\text{lb}$. Then
$3.96 \times 10^{5} \,\si{\kilogram} \times 2.2046\,\text{lb}/\si{
\kilogram} \approx 8.730 \times 10^{5}\,\text{lb}$.

\textit{Step 3, convert to slugs.} The slug is the unit of mass in
the imperial $F = ma$ system, defined so that a force of $1\,\text{lbf}$
applied to $1\,\text{slug}$ produces an acceleration of
$1\,\text{ft}/\text{s}^{2}$. Using $g_{0} = 9.80665 \,\si{\meter\per
\second\squared}$ and $1\,\text{ft} = 0.3048 \,\si{\meter}$, the slug
evaluates to $1\,\text{slug} = 14.5939 \,\si{\kilogram}$. So
$3.96 \times 10^{5} \,\si{\kilogram} \div 14.5939 \,\si{\kilogram\per
slug} \approx 2.714 \times 10^{4}\,\text{slug}$.

\textit{Step 4, assign populations.} US flight crews work in pounds
for weight-and-balance computations because the cockpit weight-and-
balance worksheet is in pounds and the manufacturer's documentation
follows. US ground engineers (structural, maintenance) work in pounds
for gross weight, but switch to slugs the moment the calculation
becomes dynamic: a load-factor or vibration analysis requires the
consistent mass unit so that $F = ma$ holds without a $g_{c}$
correction. European fuelling crews work in kilograms exclusively; the
SI mass-flow procedures and the airport fuel-uplift documentation are
in kilograms or in metric tonnes.

\textit{Step 5, sanity-check by inversion.} If $396\,\si{\tonne}$ is
the maximum take-off weight, the equivalent weight in newtons under
standard gravity is $396{,}000 \times 9.81 \approx 3.88 \times 10^{6}
\,\si{\newton}$, or about $873{,}000 \,\text{lbf}$. The pound-mass
figure in step 2 and the pound-force figure here match numerically
because $1 \,\text{lbm}$ in standard gravity weighs $1 \,\text{lbf}$;
the apparent coincidence is the source of the imperial system's
historical confusion of mass and weight.

\textit{Step 6, name the failure mode.} A calculation that uses
``pounds'' as both mass and force without distinguishing them, and
without inserting the dimensional constant $g_{c} \approx 32.2
\,\text{ft}\cdot\text{lbm}/(\text{lbf}\cdot\text{s}^{2})$, produces
results wrong by a factor of $g$. This factor-of-thirty-two error is
the classical engineering trap of the foot-pound-second system, and
it is one of the reasons SI insists on distinct units for mass
(kilogram) and force (newton).
\end{exercise}

\begin{exercise}
A torque specification is given as $40\,\si{\newton\meter}$. Convert
to foot pounds. A second specification is given as
$40\,\text{ft}\cdot\text{lbf}$. Convert to newton metres. Comment on whether
the two are interchangeable in a real-world fastener procedure.

\paragraph{Solution.}
We compute the conversions, the resulting fastener errors, and the
field practice that controls the failure mode.

\textit{Step 1, derive the conversion.} A foot-pound-force is the
torque produced by a force of $1\,\text{lbf}$ at a moment arm of
$1\,\text{ft}$. Using $1\,\text{ft} = 0.3048\,\si{\meter}$ exactly and
$1\,\text{lbf} = 4.4482216\,\si{\newton}$ (from $1\,\text{lbm} \times
9.80665\,\si{\meter\per\second\squared}$, exact), the conversion
factor is
$1\,\text{ft}\cdot\text{lbf} = 0.3048 \times 4.4482216 = 1.35582
\,\si{\newton\meter}$.
The inverse is $1\,\si{\newton\meter} = 0.737562 \,\text{ft}\cdot
\text{lbf}$.

\textit{Step 2, convert the two specifications.}
$40\,\si{\newton\meter} \times 0.7376 = 29.50 \,\text{ft}\cdot\text{
lbf}$ and $40\,\text{ft}\cdot\text{lbf} \times 1.3558 = 54.23
\,\si{\newton\meter}$.

\textit{Step 3, compute the resulting errors.} If a fastener with
specification $40\,\text{ft}\cdot\text{lbf}$ is tightened to a wrench
reading $40\,\si{\newton\meter}$, the achieved torque is
$40\,\si{\newton\meter} = 29.5\,\text{ft}\cdot\text{lbf}$, which is
$1 - 29.5/40 = 26\,\%$ below specification. If a fastener with
specification $40\,\si{\newton\meter}$ is tightened to a wrench
reading $40\,\text{ft}\cdot\text{lbf}$, the achieved torque is
$54.2\,\si{\newton\meter}$, which is $54.2/40 - 1 = 36\,\%$ above
specification.

\textit{Step 4, evaluate the consequence.} Typical preload tolerance
bands on threaded fasteners are $\pm 10$ to $\pm 25\,\%$ depending on
the joint, with the higher tolerances reserved for non-structural
applications. Neither of the unit-mismatch errors above is absorbed by
the joint's design margin. The $-26\,\%$ case under-preloads the
joint; the bolt can loosen under cyclic load and fail in fatigue at
the thread root. The $+36\,\%$ case over-preloads the joint; the bolt
can yield at the shank or strip the thread, with the failure occurring
either at installation or at the first hot thermal cycle.

\textit{Step 5, name the prevention practice.} The field practice in
mixed-fleet maintenance has three components. The wrench is labelled
with its calibrated unit system, in writing, on a tag that survives
handling. The procedure document carries explicit units on every
torque specification. One calibrated wrench is reserved per unit
system, and the technician selects the wrench by matching the
procedure's unit label. Mental conversion at the joint is not
authorised; if a conversion is required, it is performed by a second
engineer at the procurement or procedure-revision interface, in
writing, and recorded in the controlled procedure.

\textit{Step 6, record the audit trail.} The completed torque value
is logged in the work-order record with both the wrench's calibrated
unit and the specification's unit; the second engineer's signature
records the verification that the two match. The record is the
defence against the failure that the next inspector or accident
investigator will look for first.
\end{exercise}

\begin{exercise}
A medical syringe is graduated in millilitres. A drug concentration
is given as $50 \,\si{\milli\gram\per\milli\liter}$. A prescription is
written for $0.4 \,\si{\milli\gram}$, with a stated tolerance of
$\pm 5\,\%$. State the required dose volume in millilitres and the
syringe resolution that lets the dispenser meet the tolerance.
Comment on whether an ordinary $1 \,\si{\milli\liter}$ insulin
syringe is acceptable.

\paragraph{Solution.}
We compute the required volume, the tolerance on volume, the
resolution of the candidate instrument, and the clinical workflow
that closes the gap.

\textit{Step 1, compute the dose volume.} From the concentration and
the prescribed mass,
$V = m/c = 0.4 \,\si{\milli\gram} \div 50 \,\si{\milli\gram\per\milli
\liter} = 0.008 \,\si{\milli\liter} = 8 \,\si{\micro\liter}$. The
required volume is at the microlitre scale, two orders of magnitude
below the insulin syringe's nominal capacity.

\textit{Step 2, compute the volume tolerance.} The $\pm 5\,\%$
tolerance on dose translates directly to a $\pm 5\,\%$ tolerance on
volume (the concentration is fixed). The volume window is
$8 \,\si{\micro\liter} \times (1 \pm 0.05) = 7.6$ to $8.4 \,\si{
\micro\liter}$, a $\pm 0.4 \,\si{\micro\liter}$ tolerance on a
dispensed volume of $8 \,\si{\micro\liter}$.

\textit{Step 3, characterise the candidate instrument.} An ordinary
$1 \,\si{\milli\liter}$ insulin syringe has graduations every
$0.01 \,\si{\milli\liter} = 10 \,\si{\micro\liter}$. The reading
resolution is half a division, so $\pm 5 \,\si{\micro\liter}$ at
best. The instrument's stated accuracy from the manufacturer is
typically $\pm 4\,\%$ at full scale, or $\pm 40 \,\si{\micro\liter}$
at $1 \,\si{\milli\liter}$; at the low end of the scale, the relative
error grows because the absolute error stays constant.

\textit{Step 4, compare resolution to tolerance.} The instrument's
resolution of $\pm 5 \,\si{\micro\liter}$ is approximately twelve
times the required tolerance of $\pm 0.4 \,\si{\micro\liter}$. The
ordinary insulin syringe is not acceptable; even a perfect operator
cannot read the meniscus to a precision finer than the half-
division. The instrument is dispensing at the limit of its
discrimination, not at the limit of the patient's tolerance.

\textit{Step 5, name the clinical workflow.} Two paths close the gap.
The first is pre-dilution: the pharmacy dilutes the
$50 \,\si{\milli\gram\per\milli\liter}$ concentrate by a factor of
ten to $5 \,\si{\milli\gram\per\milli\liter}$, after which the
required volume is $80 \,\si{\micro\liter}$ (the tenth division of
the syringe) with tolerance $\pm 4 \,\si{\micro\liter}$ (within
reading resolution). The second is instrument substitution: a
$10 \,\si{\micro\liter}$ positive-displacement micropipette with
documented calibration delivers $\pm 0.1 \,\si{\micro\liter}$ at
$8 \,\si{\micro\liter}$ setting, comfortably inside the tolerance.

\textit{Step 6, name the lesson.} Instrument resolution must be
matched to required tolerance at the point of dispensing. The match
is made in writing on the procedure document and verified at the
point of compounding by a second pharmacist; the assumption that
``the available syringe is precise enough'' is the source of the
chronic medication-dosing errors the chapter has named. The
manufacturer's nominal precision applies at full-scale dispensing,
not at the low end of the scale where the absolute error dominates
the relative error.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Show that the standard deviation of the mean of $n$ independent
measurements, each with standard deviation $\sigma$, is
$\sigma / \sqrt{n}$. State the assumption that fails when the
measurements are not independent.

\paragraph{Solution.}
We derive the result from the definition of variance and identify the
place where independence is consumed.

\textit{Step 1, set up.} Let $X_{1}, \ldots, X_{n}$ be random variables
with $\E[X_{i}] = \mu$ and $\operatorname{Var}(X_{i}) = \sigma^{2}$
for every $i$. Define the sample mean
$\bar{X} = (1/n) \sum_{i=1}^{n} X_{i}$. We compute
$\operatorname{Var}(\bar{X})$ from the definition.

\textit{Step 2, apply linearity.} The variance of a scaled sum
satisfies
\[
\operatorname{Var}(\bar{X})
= \operatorname{Var}\!\left( \frac{1}{n} \sum_{i=1}^{n} X_{i} \right)
= \frac{1}{n^{2}} \operatorname{Var}\!\left( \sum_{i=1}^{n} X_{i}
\right).
\]
The factor $1/n^{2}$ appears because variance scales with the square
of a multiplicative constant.

\textit{Step 3, expand the variance of the sum.} The variance of a
sum can be written in terms of pairwise covariances:
\[
\operatorname{Var}\!\left( \sum_{i=1}^{n} X_{i} \right)
= \sum_{i=1}^{n} \sum_{j=1}^{n} \operatorname{Cov}(X_{i}, X_{j})
= \sum_{i=1}^{n} \operatorname{Var}(X_{i})
+ \sum_{i \ne j} \operatorname{Cov}(X_{i}, X_{j}).
\]
This identity is general; no independence assumption has been used
yet. The diagonal terms collect the variances; the off-diagonal terms
collect the covariances.

\textit{Step 4, consume independence.} If the $X_{i}$ are mutually
independent, then $\operatorname{Cov}(X_{i}, X_{j}) = 0$ for $i \ne
j$. The off-diagonal sum vanishes, and
\[
\operatorname{Var}\!\left( \sum_{i=1}^{n} X_{i} \right)
= \sum_{i=1}^{n} \sigma^{2}
= n \sigma^{2}.
\]
Combining with step 2,
$\operatorname{Var}(\bar{X}) = (1/n^{2}) \cdot n \sigma^{2} =
\sigma^{2}/n$. Taking the square root,
$\operatorname{SD}(\bar{X}) = \sigma / \sqrt{n}$.

\textit{Step 5, name the failure mode.} The independence assumption is
consumed exactly once, in step 4. If the measurements are correlated,
the off-diagonal sum does not vanish. In the worst case (all
measurements perfectly correlated), every covariance equals
$\sigma^{2}$, and
$\operatorname{Var}(\sum X_{i}) = n^{2} \sigma^{2}$, so
$\operatorname{Var}(\bar{X}) = \sigma^{2}$: averaging gains nothing.
In the intermediate case (correlation $\rho > 0$ between pairs),
$\operatorname{Var}(\bar{X}) = \sigma^{2} [1/n + (n-1)\rho/n]
\to \rho \sigma^{2}$ as $n \to \infty$; averaging hits a floor at
$\sigma \sqrt{\rho}$ rather than zero.

\textit{Step 6, list the engineering violations.} Sources that
violate independence in practice include: instrument drift (every
later reading is correlated with the earlier ones through the slow
trend); common-mode environmental forcing (every reading on a hot
afternoon shares the same temperature bias); shared calibration error
(every reading uses the same reference, so a calibration mistake
propagates identically); a shared operator (every reading carries the
same systematic operator bias); a shared computational step (every
reading is processed through the same buggy script). Each of these
turns the $1/\sqrt{n}$ promise into a slower decay or a floor; the
reader who reports the standard deviation of the mean as the
measurement uncertainty without checking for correlation has hidden
the dominant error.
\end{exercise}

\begin{exercise}[\textit{deferred to Chapter 4}]
Show that, in the small-error limit, the relative uncertainty of a
product $z = xy$ satisfies
\[
\left(\frac{u(z)}{z}\right)^{2}
= \left(\frac{u(x)}{x}\right)^{2}
+ \left(\frac{u(y)}{y}\right)^{2}
\]
when $x$ and $y$ are uncorrelated. Identify a domain in which this
formula systematically underestimates the uncertainty. Chapter 4 of
this volume derives the general propagation rule; readers can return
here after that chapter.
\end{exercise}

\begin{exercise}[\textit{deferred to Chapter 4}]
Derive the propagation rule for $z = x + y$ when $x$ and $y$ have
correlation $\rho$. State the result and identify the case in which
the uncertainty of $z$ is smaller than either $u(x)$ or $u(y)$. Read
chapter 4's covariance section before attempting.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the total mass of water in a typical human adult. State
your assumptions, your method, and a factor-of-three confidence
range. Then look up the published figure and report the ratio of
your estimate to the published value.

\paragraph{Reference estimate.} A typical adult mass is
$\sim 70 \,\si{\kilogram}$. Biological tissue is approximately
$50$-$70\,\%$ water by mass, varying with body composition (lean
tissue $\sim 70\,\%$, adipose $\sim 20\,\%$, bone $\sim 30\,\%$).
Taking a working average of $60\,\%$ gives an estimate of
$\sim 42 \,\si{\kilogram}$, with a factor-of-three bracket of
$14$ to $130 \,\si{\kilogram}$ (the bracket is generous; the actual
spread for healthy adults is closer to $35$-$50 \,\si{\kilogram}$).
The published reference value from the ICRP reference adult is
$\sim 42 \,\si{\kilogram}$ for the $70 \,\si{\kilogram}$ reference
male \cite{std:icrp-89-2002}. The ratio of estimate to reference is
close to unity; the estimation habit's value here is the explicit
recording of the assumption (water fraction) that determined the
answer.
\end{exercise}

\begin{exercise}
Estimate the mass of the Eiffel Tower. State assumptions and
confidence range; do not look up the answer until you have written
your estimate.

\paragraph{Reference estimate.} Approach by enclosing volume. The
tower is approximately $324 \,\si{\meter}$ tall on a base of
roughly $125 \times 125 \,\si{\meter}$; the enclosed pyramidal
volume is $\sim (1/3) \times 125^2 \times 324 \approx 1.7 \times
10^6 \,\si{\meter\cubed}$. The tower is wrought iron lattice with
substantial void fraction; a working guess of $\sim 0.1\,\%$ iron
by volume gives $\sim 1700 \,\si{\meter\cubed}$ of iron at density
$\sim 7800 \,\si{\kilogram\per\meter\cubed}$, yielding an estimate
of $\sim 1.3 \times 10^7 \,\si{\kilogram} = 13{,}000 \,\si{\tonne}$.
Factor-of-three bracket: $4{,}000$ to $40{,}000 \,\si{\tonne}$. The
published value for the iron structure is $\sim 7{,}300 \,\si{\tonne}$
\cite{web:eiffel-tower-official}; with paint and accumulated
modifications the total is closer to $\sim 10{,}000 \,\si{\tonne}$.
The estimate brackets the published value; the dominant source of
error in the calculation was the void-fraction assumption.
\end{exercise}

\begin{exercise}
Estimate the energy stored in a fully charged AA alkaline battery.
Express in (a) joules, (b) watt hours. Compare to the energy stored
in a fully charged smartphone battery and identify which is larger
and by what factor.

\paragraph{Reference estimate.} An AA alkaline cell delivers nominal
voltage $\sim 1.5 \,\si{\volt}$ and capacity $\sim 2000 \,\si{\milli
\ampere\hour} = 2 \,\si{\ampere\hour}$. Energy: $E = V \times Q
\approx 1.5 \times 2 = 3 \,\si{\watt\hour} \approx 10{,}800 \,\si{
\joule}$. A typical smartphone battery is $\sim 3.7 \,\si{\volt}$
lithium-ion at $\sim 4000 \,\si{\milli\ampere\hour}$, giving
$\sim 15 \,\si{\watt\hour} \approx 54{,}000 \,\si{\joule}$. The
smartphone stores $\sim 5\times$ the energy of one AA cell. Sanity
check: a phone runs for roughly $24 \,\si{\hour}$ at $\sim 0.6 \,\si{
\watt}$ idle, consistent with $\sim 15 \,\si{\watt\hour}$ capacity.
The estimation habit catches sign errors and order-of-magnitude
mistakes (e.g., confusing watt-hours and milliampere-hours, or
omitting the voltage in the energy product).
\end{exercise}

\begin{exercise}
Estimate how many sensor readings your phone records during one
minute while sitting still with the screen on. Count only sensors you
can name from settings, documentation, or an installed sensor logger.
State what you excluded and why.

\paragraph{Reference estimate.} A modern smartphone runs the
following sensors at the listed nominal rates: accelerometer
($\sim 50$-$200 \,\si{\hertz}$); gyroscope ($\sim 50$-$200 \,\si{
\hertz}$); magnetometer ($\sim 10$-$50 \,\si{\hertz}$); GPS
($\sim 1 \,\si{\hertz}$); ambient light ($\sim 5 \,\si{\hertz}$);
proximity ($\sim 5 \,\si{\hertz}$); pressure/barometer ($\sim 1-10
\,\si{\hertz}$); touch-screen ($\sim 60$-$120 \,\si{\hertz}$); audio
microphone ($\sim 44{,}100 \,\si{\hertz}$ if recording); camera (only
if open). Summing nominal rates excluding microphone and camera:
$\sim 200 \,\si{\hertz}$ aggregate. Per minute: $\sim 12{,}000$
samples. Including microphone: $\sim 2.6 \times 10^6$ samples per
minute. The exact answer depends on which sensors are active for
which app; the reader is invited to install a sensor logger (e.g.,
\texttt{phyphox}) and confirm. The estimate's value is recognising
that a phone in routine operation produces measurement data at
rates comparable to a small scientific instrument.
\end{exercise}

\begin{exercise}
Use a cited global cement production figure from USGS, the Global
Cement and Concrete Association, or another named dataset; state the
year and source. Compute the mass flow rate in tonnes per second and
comment on what that flow rate implies for the engineering
disciplines that use cement.

\paragraph{Reference estimate.} Annual global cement production was
$\sim 4.1 \times 10^9 \,\si{\tonne}$ (current as of 2023, USGS
Mineral Commodity Summaries \cite{data:usgs-mcs-cement-2024}). The
mass flow rate is $4.1 \times 10^9 \,\text{tonnes}/\text{yr}
\div (3.15 \times 10^7 \,\text{s}/\text{yr}) \approx
130 \,\si{\tonne\per\second}$. The dataset
\texttt{data/cement\_production\_usgs.csv} (co-located with this
chapter) carries the year-by-year values from 2010 to 2023. The
implication for engineering disciplines: cement production at
$130 \,\si{\tonne\per\second}$ implies that the global civil-and-
construction engineering profession is operating at a continuous
material-flow scale comparable to the largest single industrial
process humans have built, with corresponding implications for
quality-control (every tonne must meet a specification), supply-
chain (the kiln-to-site logistics is itself a major engineering
domain), and emissions (cement is $\sim 8\,\%$ of global CO$_2$).
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
In a language of the reader's choice, simulate $10^{4}$ independent
measurements of a quantity drawn from $\mathcal{N}(\mu, \sigma)$ with
$\mu = 100$ and $\sigma = 5$. For sample sizes $n = 1, 4, 9, 16, 25,
\dots, 100$, plot the standard deviation of the mean as a function
of $n$. Confirm the $1/\sqrt{n}$ scaling and report the agreement
to three significant figures at $n = 100$.

\paragraph{Solution sketch.} The expected standard deviation of the
mean at $n=100$ is $5/\sqrt{100} = 0.500$. A representative
simulation (see \texttt{code/monte\_carlo\_sigma.py}) returns
$0.498 \pm 0.005$ across $10^4$ Monte Carlo trials, consistent with
$1/\sqrt{n}$ to three significant figures. The plot is straight on a
log-log axis with slope $-1/2$. The deviation from the asymptotic
$1/\sqrt{n}$ at small $n$ is the finite-sample correction; the
chi-squared distribution of the sample variance produces the visible
scatter around the analytic curve. Readers comparing implementations
across languages should expect agreement to the fourth significant
figure with $10^4$ trials; agreement to better than that requires
larger Monte Carlo runs.
\end{exercise}

\begin{exercise}
Modify the previous exercise so that each measurement has a
systematic offset of $\Delta = 2$ in addition to the random error.
Plot the apparent standard deviation of the mean and the apparent
mean against $n$. Demonstrate, with a sentence, why the systematic
component does not vanish with averaging.

\paragraph{Solution.}
We compute each curve, distinguish bias from dispersion, and connect
the result to the GUM combined-uncertainty machinery.

\textit{Step 1, set up the model.} Each measurement is
$X_{i} = \mu + \Delta + \varepsilon_{i}$ where $\mu = 100$ is the
true value, $\Delta = 2$ is a constant systematic offset, and
$\varepsilon_{i} \sim \mathcal{N}(0, \sigma)$ with $\sigma = 5$ is
the random component. The two effects enter additively.

\textit{Step 2, compute the apparent-mean curve.} For sample size
$n$, the apparent mean is
$\bar{X} = (1/n) \sum_{i} X_{i} = (\mu + \Delta) + (1/n)\sum_{i}
\varepsilon_{i}$.
The expected value is $\E[\bar{X}] = \mu + \Delta = 102$ for every
$n$. As $n \to \infty$, the random term vanishes by the law of
large numbers, so $\bar{X} \to 102$. The plot of $\bar{X}$ vs $n$
approaches a horizontal line at $102$, not at $\mu = 100$.

\textit{Step 3, compute the apparent-standard-deviation-of-mean
curve.} The variance of $\bar{X}$ is
$\operatorname{Var}(\bar{X}) = \operatorname{Var}((1/n)\sum_{i}
\varepsilon_{i}) = \sigma^{2}/n$. The standard deviation is
$\sigma/\sqrt{n} = 5/\sqrt{n}$. The systematic component $\Delta$
does not appear in the variance: it is a constant, and the variance
of a random variable plus a constant is the variance of the random
variable. The plot of $\operatorname{SD}(\bar{X})$ vs $n$ on
log-log axes is a straight line of slope $-1/2$, passing through
$5$ at $n = 1$ and $0.5$ at $n = 100$.

\textit{Step 4, distinguish bias from dispersion.} Bias and
dispersion are independent components of the total error. Averaging
reduces dispersion as $1/\sqrt{n}$ but does not affect bias at any
$n$. The reported measurement converges with arbitrarily small
spread on the wrong value. An observer who tracks only the
dispersion (the standard deviation of the mean) sees the dispersion
shrink and concludes that the measurement is becoming more
accurate; the bias remains, hidden from the dispersion-based
analysis.

\textit{Step 5, name the working-protocol failure mode.} A
measurement protocol that reports the standard deviation of the
mean as the measurement uncertainty has hidden the bias entirely.
The reported uncertainty is $\sigma/\sqrt{n}$; the actual error of
the reported value is at least $|\Delta|$ for every $n$ and may be
much larger if there are multiple unrecognised systematic
contributors. A reader using the reported uncertainty for
downstream propagation is being misled.

\textit{Step 6, name the GUM response.} The Guide to the Expression
of Uncertainty in Measurement (GUM) requires the engineer to
evaluate both Type A (random, derived from the statistical analysis
of repeated measurements) and Type B (systematic, derived from
calibration records, instrument specifications, and known physical
effects) components. The combined standard uncertainty is the
quadrature sum: $u_{c} = \sqrt{u_{A}^{2} + u_{B}^{2}}$. In the
present case, $u_{A} = \sigma/\sqrt{n}$ from the standard deviation
of the mean, and $u_{B} = |\Delta|/\sqrt{3} \approx 1.15$ if
$\Delta$ is treated as a uniformly distributed systematic bias
within $\pm |\Delta|$. The reader who reports $u_{c}$ rather than
$u_{A}$ alone is reporting an honest measurement; the reader who
reports $u_{A}$ alone is reporting a precision claim that the
accuracy does not support. See
\texttt{code/monte\_carlo\_sigma.py} for the modified simulation.
\end{exercise}

\begin{exercise}
Generate $1000$ pairs $(x, y)$ from a bivariate normal with
$\sigma_{x} = \sigma_{y} = 1$ and correlation $\rho \in
\{-0.9, 0, 0.9\}$. Compute the sample standard deviation of $z = x +
y$ in each case and compare to the analytical result.

\paragraph{Solution.}
We derive the analytical variance, compute it at each correlation,
verify against simulation, and connect the negative-correlation case
to differential-measurement instrumentation.

\textit{Step 1, derive the analytical variance.} For a sum
$z = x + y$ of two random variables with variances $\sigma_{x}^{2}$,
$\sigma_{y}^{2}$ and correlation $\rho$,
\[
\operatorname{Var}(z)
= \operatorname{Var}(x) + \operatorname{Var}(y) + 2 \operatorname{Cov}
(x, y)
= \sigma_{x}^{2} + \sigma_{y}^{2} + 2 \rho \sigma_{x} \sigma_{y}.
\]
With $\sigma_{x} = \sigma_{y} = 1$, this simplifies to
$\operatorname{Var}(z) = 2(1 + \rho)$.

\textit{Step 2, compute at each correlation.}
For $\rho = -0.9$, $\operatorname{Var}(z) = 2 \times 0.1 = 0.2$, so
$\sigma_{z} = 0.447$.
For $\rho = 0$, $\operatorname{Var}(z) = 2$, $\sigma_{z} = \sqrt{2}
\approx 1.414$.
For $\rho = 0.9$, $\operatorname{Var}(z) = 2 \times 1.9 = 3.8$,
$\sigma_{z} = 1.949$.

\textit{Step 3, verify against simulation.} The simulation
(see \texttt{code/bivariate\_propagation.py}) draws $1000$ pairs
$(x, y)$ from a bivariate normal at each correlation value and
computes the sample standard deviation of $z = x + y$. Representative
output: $0.446$, $1.412$, $1.948$ for the three correlations
respectively. The agreement to three significant figures is the
expected $1/\sqrt{1000} \approx 3\,\%$ statistical uncertainty in
the sample standard deviation; doubling the sample size halves the
discrepancy.

\textit{Step 4, identify the engineering insight.} The negative-
correlation case is the working insight: when two error sources are
anti-correlated, the sum is more precise than either term, because
the errors partially cancel. With $\sigma_{x} = \sigma_{y} = 1$ and
$\rho = -0.9$, the sum has $\sigma_{z} = 0.45$, less than half of
either input's standard deviation. As $\rho \to -1$ the sum's
variance approaches zero; the perfect-anti-correlation limit is a
noiseless difference between two equally noisy inputs.

\textit{Step 5, connect to differential-measurement instrumentation.}
The negative-correlation mechanism is the structural reason that
differential measurement reduces noise. The Wheatstone bridge
measures the small difference between two large resistances; the
ratiometric ADC compares the measured signal to a reference that
shares the same supply-noise; the common-mode-rejection amplifier
subtracts the common-mode component between two input lines. In each
case, the correlated noise component cancels in the difference, and
the residual is the small uncorrelated component that carries the
actual measurement.

\textit{Step 6, name the design pattern.} The pattern generalises:
whenever two error sources share a common cause (temperature
drift, power-supply ripple, mechanical vibration, common reference),
the engineer can subtract them by routing the measurement and a
matched dummy through the same noise environment. The signal
appears in the difference; the noise cancels. The instrumentation
amplifier, the bridge configuration, and the differential ADC are
three implementations of the same principle, all licensed by the
$\sigma_{z}^{2} = \sigma_{x}^{2} + \sigma_{y}^{2} - 2\rho \sigma_{x}
\sigma_{y}$ identity (with sign-flipped second variable, i.e., for
the difference $z = x - y$ the variance is the same expression with
$-2\rho$, so positive correlation reduces the difference's variance).
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a kitchen experiment that measures the local acceleration due
to gravity, $g$, to within $5\,\%$ using only a tape measure, a
stopwatch, and one household object the reader chooses. State your
method, your expected error sources, and the dominant one.

\paragraph{Rubric.} A credible design names: the method (pendulum
period $T = 2\pi\sqrt{L/g}$ with a $1$-$2 \,\si{\meter}$ pendulum,
or free-fall timing from a measured drop height); the dominant
error sources (for the pendulum: timing reaction-time of order
$0.1$-$0.3 \,\si{\second}$ on a period of $2$-$3 \,\si{\second}$,
hence the engineering response of timing $10$ periods and dividing;
length-measurement of order $\pm 5 \,\si{\milli\meter}$ on
$\sim 1 \,\si{\meter}$, hence $\pm 0.5\,\%$); the dominant
contributor under that protocol (the timing of single periods if
not averaged; the length measurement if averaged). A design that
achieves $5\,\%$ accuracy without naming reaction time has not
identified the dominant error. Rubric scoring: identification of
the method (1), error-source enumeration (1), dominant-error
identification with quantitative reasoning (2).
\end{exercise}

\begin{exercise}
Design a household procedure for verifying the calibration of a
kitchen scale to within $2\,\%$. Use one of three reference classes:
a calibration mass with a stated tolerance, coins whose official mass
is published for the reader's country, or a sealed package with a
labelled net mass and stated packaging tolerance. State the
reference and its uncertainty, and explain why this is a check, not a
traceable calibration.

\paragraph{Rubric.} A credible procedure names: the chosen reference
class (coins are widely available and have published masses; see
\texttt{data/coin\_masses.csv}); the reference's stated uncertainty
(coin masses are specified by the issuing mint typically to
$\pm 0.5\,\%$ for circulation coins, tighter for collector coins);
the verification protocol (weigh the reference at a known mass and
record the deviation; repeat with multiple references spanning the
scale's range); the criterion for $2\,\%$ accuracy (the deviation
must be below $2\,\%$ at every reference mass). The distinction from
a traceable calibration is the absence of a documented traceability
chain to a national or international standard: the coin reference is
not itself calibrated against a NIST-traceable standard with a
calibration certificate; the user has no record of when the coin
was last verified; the in-circulation wear can shift the mass beyond
the mint specification. A traceable calibration uses certified
reference masses with documented traceability and a calibration
certificate from an accredited laboratory.
\end{exercise}

\begin{exercise}
Design a measurement plan to determine whether a household
refrigerator's temperature is meeting its setpoint. Specify
instruments, sampling rate, sampling locations within the
refrigerator, duration, the analysis, and the criterion for
``meeting the setpoint.'' Comment on what would change for a
clinical-grade vaccine refrigerator.

\paragraph{Rubric.} A credible plan specifies: instrument (a logging
thermometer with $\pm 0.5 \,\degC$ accuracy is adequate for a
household refrigerator setpoint of typically $4 \,\degC$; a Type-K
thermocouple data logger or a USB-temperature probe both work);
sampling rate ($\geq 1 \,\text{sample/minute}$, since the cycle
period of a household refrigerator's compressor is typically
$15$-$60 \,\text{minutes}$); locations (top, middle, bottom shelf
plus door-shelf and crisper drawer; the temperature gradient
between these is typically $1$-$3 \,\degC$); duration ($\geq 24
\,\text{hours}$ to capture multiple compressor cycles and to span
the day-night ambient cycle); analysis (compute the mean, the
$\pm 1\sigma$ spread, the minimum and maximum at each location);
criterion (the mean temperature at the storage area is within
$\pm 1 \,\degC$ of setpoint and the maximum stays below $\sim 7
\,\degC$ for food safety). For a clinical-grade vaccine
refrigerator: the requirements are codified in WHO PQS standards
\cite{web:who-pqs-refrigerators-2024}; the temperature range is
specified ($2$-$8 \,\degC$); the monitoring is continuous with
alarms; the calibration is traceable; the validation is documented.
The shift from household to clinical is from informal check to
regulated process.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A torque wrench is dropped on a hard floor. The user resumes work
without recalibrating. State three plausible failure modes the wrench
may now exhibit and propose a household-feasible verification for
each.

\paragraph{Solution.}
We name three plausible mechanisms, design a household-feasible
verification for each, and state the working-professional response.

\textit{Step 1, list the failure modes by sub-system.} A click-type
torque wrench has three load-bearing sub-systems: the calibrated
spring-and-cam that triggers the click; the detent surfaces that
release at the set torque; and the scale-and-handle assembly that
sets the trigger threshold. A drop can damage any of the three.

\textit{Step 2, name failure mode A (offset).} The spring's preload
or the cam geometry shifts, so the click occurs at a different torque
than the scale indicates. The result is a systematic offset between
set torque and applied torque, the same magnitude for every reading.
Household verification: clamp the wrench horizontally in a vice with
the head fixed. Hang a known weight $W$ on the handle at a known
lever arm $L$. The applied torque is $W L$. Set the wrench to $W L$,
release the load slowly until the click triggers, and compare to the
computed $W L$. Repeat at three set torques (e.g.,
$20$, $40$, $60 \,\si{\newton\meter}$) to map the offset across the
range.

\textit{Step 3, name failure mode B (precision).} The detent surfaces
or the pivot bushings are damaged, producing an inconsistent release
torque on repeated trials. The result is a spread of click torques
around the set value, not a constant offset. Household verification:
repeat the same set torque ten times against the same reference
(weight on lever arm, as above). Compute the standard deviation. A
serviceable wrench shows a spread of $\pm 2$-$4\,\%$ of full scale;
a damaged wrench shows $\pm 10\,\%$ or larger, with audible roughness
in the click.

\textit{Step 4, name failure mode C (display).} The scale display or
the handle's locking ring is mechanically misaligned, so the value
read off the scale does not match the value set on the spring.
Household verification: with the wrench horizontal, set the scale to
a known value (e.g., $40 \,\si{\newton\meter}$), and use a magnifying
glass to read the indicator. If the indicator is off-axis or the
locking ring is bent, the displayed value differs from the indicated
value by a constant amount. The check is visual and does not require
a load test.

\textit{Step 5, distinguish failure modes A and B from results.} The
results of step 2 and step 3 distinguish the mechanisms. A constant
offset across set torques implies mode A; a varying spread at fixed
set torque implies mode B; both can be present simultaneously and the
analysis remains valid because the two contribute additively (the
mean of step 3's ten trials is the offset of mode A; the spread is
mode B).

\textit{Step 6, state the working response.} The household
verification is a diagnostic for whether the wrench is fit for
unsupervised use on non-critical fasteners; it is not a substitute
for accredited calibration. A working professional removes a dropped
torque wrench from service immediately, tags it ``do not use,'' and
sends it to an accredited calibration laboratory. The laboratory
returns a calibration certificate stating the wrench's accuracy and
its calibration date; the certificate is the documentation that
survives the next failure investigation. The household check is for
the home shop; the professional shop has a recall procedure for
dropped wrenches.
\end{exercise}

\begin{exercise}
A car's fuel-economy display reports $5.0\,\text{L}/100\,\text{km}$.
The driver fills the tank, drives a measured distance, fills again,
computes the actual consumption, and finds
$5.6\,\text{L}/100\,\text{km}$. State at least three sources of the
discrepancy, then name the first measurement you would repeat before
blaming the dashboard algorithm.

\paragraph{Solution.}
We compute the magnitude of the discrepancy, list candidate sources
with quantitative weights, and name the first measurement to repeat.

\textit{Step 1, quantify the discrepancy.} The dashboard reports
$5.0 \,\text{L}/100\,\text{km}$; the measured value is
$5.6 \,\text{L}/100\,\text{km}$. The ratio is $5.6/5.0 = 1.12$, or
$+12\,\%$ relative to the dashboard. This is a non-trivial gap; a
$\pm 5\,\%$ tolerance for a dashboard consumption display would be
optimistic, but a $\pm 12\,\%$ gap warrants investigation.

\textit{Step 2, list candidate sources.} Three independent
measurements determine the computed consumption: the distance, the
fuel volume on the first fill, and the fuel volume on the second
fill. Each is subject to its own error sources.

\textit{Step 3, weight the odometer source.} The odometer infers
distance from wheel rotation count and an assumed rolling radius. The
rolling radius depends on tyre size, wear, pressure, and temperature;
typical odometer accuracy is $\pm 3$-$5\,\%$, with worn or under-
inflated tyres reading the distance high (because the actual radius
is less than the assumed radius, so each wheel rotation covers less
distance than the odometer counts). A $+5\,\%$ odometer bias would
overstate distance, understate consumption-per-distance, and make the
dashboard look better than the truth. The direction of the typical
bias is consistent with the observation.

\textit{Step 4, weight the pump source.} A fuel pump's auto-shutoff
triggers when fuel rises into a vent passage near the nozzle tip; the
trigger level depends on the angle of the nozzle, the splash
behaviour at the tank's filler neck, and the dispensing rate at the
moment of shutoff. Per-fill variation is typically $\pm 0.5$ to
$\pm 1 \,\si{\liter}$ on a $\sim 50 \,\si{\liter}$ tank, which is
$\pm 1$-$2\,\%$ per fill, or $\pm 2$-$4\,\%$ on the difference of
two fills. The two-fill protocol amplifies pump noise.

\textit{Step 5, weight the dashboard model.} The dashboard computes
consumption from the engine's fuel-injector pulse-width and the
air-mass-flow sensor, with the air-fuel ratio assumed from the
engine-management calibration. This is a model of fuel flow, not a
direct measurement of the fuel leaving the tank. Typical accuracy is
$\pm 5$-$10\,\%$; deliberate engine-management bias toward optimistic
consumption reports is not unknown. The dashboard model is therefore
the largest single candidate.

\textit{Step 6, name the first measurement to repeat.} The pump-to-
pump fuel quantity. Fill the tank from the same pump at the same
station on two consecutive days with no driving between the fills; if
the dispensed quantity differs by less than $\pm 0.5 \,\si{\liter}$,
the fuel-volume measurement is reproducible and the dashboard model
is the likely source of the $+12\,\%$ gap. If the dispensed quantity
differs by more, the pump auto-shutoff dominates the measurement noise
and the comparison itself is not reproducible at the required level.
The first measurement to repeat is therefore the one whose
reproducibility we cannot assume, not the one whose accuracy we wish
to verify.
\end{exercise}

\begin{exercise}
A laboratory pipette consistently delivers $98 \,\si{\micro\liter}$
when set to $100 \,\si{\micro\liter}$, with a spread of
$\pm 1 \,\si{\micro\liter}$. The pipette is used in a clinical assay
that has reported drift in its results over the last six months. State
whether the pipette is the cause and what evidence would distinguish
it from other candidates.

\paragraph{Solution.}
We characterise the pipette, distinguish offset from drift, list
candidate sources, and design the distinguishing measurements.

\textit{Step 1, characterise the pipette.} The pipette has a
systematic offset of $-2\,\%$ (delivers $98 \,\si{\micro\liter}$ when
set to $100 \,\si{\micro\liter}$) and a random spread of
$\pm 1\,\%$. The instrument is precise (small spread) and inaccurate
(non-zero bias). Both numbers are quoted relative to the set value.

\textit{Step 2, distinguish offset from drift.} A constant bias
produces a constant offset in every assay result, not a time-varying
trend. The assay has reported \emph{drift}, which is a change over
time. The pipette is the proximate cause of drift only if the bias
itself has changed over the six-month period.

\textit{Step 3, list candidate drift sources with engineering
mechanism.} (i) Reagent degradation: assay reagents lose activity
with storage time and temperature; a slow drift in reagent
concentration produces a slow drift in measured result. (ii)
Reference-standard drift: calibration of the assay's reference
standard against a higher-tier reference may have drifted; drift in
the reference appears as drift in every measurement made against it.
(iii) Reading instrument: a spectrophotometer's lamp output decreases
with hours of use, biasing the absorbance reading. (iv) Operator
change: a new operator with different pipetting technique (angle,
pre-wet, dispense speed) produces a step change masquerading as drift
across the rota. (v) Pipette wear: the pipette's piston seal or tip
fitting may have degraded, growing the bias from $-2\,\%$ at
calibration to $-4\,\%$ or worse by month six.

\textit{Step 4, design the distinguishing measurements.} Three
measurements separate the candidates.

First, pipette calibration history: the pipette's quarterly
calibration record should show whether the $-2\,\%$ bias was constant
across the six months. If the record shows a worsening trend, the
pipette is a candidate. If the record shows constant bias, the
pipette is exonerated.

Second, control-sample trending: run a known control sample weekly
across the next four weeks with the suspect pipette and an
independently calibrated second pipette. If both pipettes track the
same drift, the drift is in reagents, reference, or instrument
downstream. If the two pipettes disagree and the disagreement matches
the bias differential, the pipette is the source.

Third, reading-instrument validation: run the same control sample
through the spectrophotometer with a freshly recalibrated
absorbance reference. If the result matches the reagent-and-reference
prediction, the spectrophotometer is in tolerance; if it disagrees,
the spectrophotometer is drifting.

\textit{Step 5, weigh the candidates.} The pipette is unlikely to be
the dominant cause, because its measured bias is constant and the
assay drift is time-varying. Reagent degradation and reference-
standard drift are the two most common sources of clinical-assay
drift over a six-month window; the literature on clinical-laboratory
quality control treats reagent stability as the most frequent
suspect.

\textit{Step 6, name the working practice.} The clinical laboratory
runs a daily control sample with known expected value, plots the
result against a control chart, and flags the assay for investigation
when the chart shows a Western-Electric rule violation (e.g., two
consecutive points outside $\pm 2\sigma$, or seven consecutive points
on one side of the mean). The control chart, not the post-hoc
investigation, is the engineering response to assay drift; the
present diagnosis is the corrective action after the chart has
flagged.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Read the executive summary of the Mars Climate Orbiter mishap report
\cite{acc:nasa-mco-mib}. Identify (a) the named root cause as the
report states it, (b) at least one contributing factor the report
identifies but does not call root, (c) one organisational cause
implicit in the report's narrative.

\paragraph{Solution.}
We work through the report's three layers: stated root cause,
enumerated contributing factors, and implicit organisational cause.

\textit{Step 1, locate the root cause statement.} The Mishap
Investigation Board report names the root cause as the failure to use
metric units in the coding of a specific ground software file
(SM\_FORCES) used in the trajectory models. The file produced impulse
values in pound-force seconds; the navigation software consumed the
values as newton-seconds. The conversion factor of
$1 \,\text{lbf}\cdot\text{s} = 4.45 \,\si{\newton\second}$ was missing
from the interface specification and from the data exchange.

\textit{Step 2, quantify the consequence.} Over the eleven-month
cruise to Mars, the small-forces accumulated to produce a trajectory-
modelled position that was approximately $170 \,\si{\kilo\meter}$
closer to the planet than the actual spacecraft was. The Mars
Orbital Insertion was computed for the modelled trajectory; the
actual spacecraft entered the Martian atmosphere at an altitude of
approximately $57 \,\si{\kilo\meter}$, below the $80 \,\si{\kilo
\meter}$ threshold at which the heat shield-less orbiter could
survive. The orbiter was lost.

\textit{Step 3, list contributing factors the report enumerates.} The
report identifies several. Inadequate verification of the navigation
model against in-flight data: the small-forces accumulation was
visible in the telemetry and was queried by the navigation team, but
the queries were not escalated to a system-level review. Sparse
small-forces telemetry: the data exchange between contractor and
operator was not configured to make the unit-system visible to either
party. Under-staffing of the navigation team: the small-forces
investigation was deferred under workload pressure. Absence of
independent cross-checks at the operations interface: no party was
charged with verifying the unit consistency of the exchanged data.

\textit{Step 4, identify the implicit organisational cause.} The
report does not name the organisational fragmentation as a root cause
directly, but its narrative makes the cause visible. The ground
software was developed by Lockheed Martin; the navigation model was
operated by JPL; the data exchange was specified in a document that
treated unit consistency as a downstream concern. No single party was
accountable for end-to-end unit consistency, and the unit-system
choice (imperial on the contractor side, SI on the operator side) was
not the subject of a defined interface specification. The
fragmentation made the failure invisible to each party at the time of
the relevant decision.

\textit{Step 5, distinguish the report's recommendations.} The report
recommends interface-document standardisation, independent navigation
verification, and increased small-forces telemetry resolution. These
recommendations address the technical layers of failure. The
organisational recommendation (single end-to-end accountability for
mission-critical data exchanges) is implicit in the report's
discussion of contractor-operator interfaces but is not stated as a
recommendation at the same level.

\textit{Step 6, connect to the chapter's discipline.} The
chapter's seven-component measurement habit (dated entry, instrument
and calibration, operator, conditions, provenance, signature, reading)
would have made the unit-system choice visible at the
contractor-operator interface; the missing component in the MCO data
exchange was the explicit unit declaration on every data field. The
discipline does not prevent the organisational fragmentation, but it
does make the technical failure visible at the point where the
fragmentation produces it.
\end{exercise}

\begin{exercise}
The Air Canada Flight 143 incident involved a unit conversion error,
an unserviceable fuel quantity processor, and procedural shortcuts
\cite{acc:lockwood-gimli1985}. State, in one sentence each, what
each of the three contributed. Then state which of the three a
measurement discipline of the kind described in section 1.5 would
have caught on its own and which would still require a procedural
change.

\paragraph{Solution.}
We work each contribution, compute the resulting error, and assign
the corrective action to its appropriate engineering layer.

\textit{Step 1, name contribution one and compute the error.} The
unit conversion error used $1.77 \,\text{lb/L}$ where the calculation
required the metric specific gravity of approximately
$0.80 \,\si{\kilogram\per\liter}$. The two numbers differ by the
pound-to-kilogram factor of $2.205$. The required fuel mass was
under-stated by this factor: instead of loading the mass needed for
the planned flight, the fuelling crew loaded the mass that would
satisfy the calculation if pounds were the unit. The aircraft
departed with approximately $45\,\%$ of the fuel required for the
planned distance.

\textit{Step 2, name contribution two and compute the missing
defence.} The fuel quantity processor (FQP) was the onboard system
that read fuel quantity directly from tank sensors and displayed it
in the cockpit. It was unserviceable on Flight 143. With the FQP
working, the crew would have seen at start-up that the displayed
fuel mass was approximately half the expected value, and would have
queried the discrepancy at the gate before pushback. With the FQP
inoperative, the cockpit display was blank and the manual
calculation became the sole source of the fuel-mass figure.

\textit{Step 3, name contribution three and the procedural shortcut.}
The Minimum Equipment List (MEL) for the Boeing 767 at the time
permitted dispatch with the FQP inoperative under specific
conditions, including a manual dip-stick measurement of fuel quantity
at each fuel uplift. The manual procedure was performed; the dip-
stick reading in litres was converted to mass using the wrong
conversion factor; the resulting mass figure was carried forward
without an independent check. The procedural authorisation removed
the institutional barrier (no dispatch without FQP) that would have
caught the manual-conversion error.

\textit{Step 4, decompose the engineering layers.} Three layers
contributed: the measurement layer (manual conversion arithmetic),
the instrumentation layer (FQP unserviceable), and the procedural
layer (MEL authorising dispatch). The accident was the simultaneous
failure of all three; any one of them held would have prevented the
fuel exhaustion at altitude.

\textit{Step 5, assign correctives by layer.} The seven-component
measurement discipline of section 1.5 catches contribution one. Each
component of the calculation log carries its unit explicitly; the
conversion ratio carries its dimensional check; the calculation
result carries its unit. A second crew member counter-signs the
calculation against the source data. The discipline would have
flagged the dimensional inconsistency at the gate, well before
pushback.

The measurement discipline also catches contribution two indirectly,
by recording the instrument's calibration and serviceability status
on every entry that uses its reading; in this case the entry would
read ``FQP inoperative; mass derived from dip-stick measurement at
gate.'' The recording makes the dependency on the manual path
visible.

The procedural authorisation in contribution three is institutional.
The corrective is a change to the MEL or to the dispatch procedure;
the measurement discipline alone does not address it. Post-accident
Transport Canada revised the dispatch procedure to require a second
independent fuel-mass calculation under the manual path.

\textit{Step 6, name the chapter's broader lesson.} A measurement
discipline catches measurement errors. It does not catch institutional
errors. The seven-component habit makes the measurement layer
auditable; the audit reveals where the institutional layer has
abandoned its defences. The lesson of Gimli is that measurement
discipline is necessary but not sufficient: the institutional layer
must be designed to preserve defences when the measurement layer
reveals an unusual dependency, and the discipline is the mechanism
that makes the dependency visible.
\end{exercise}

\begin{exercise}
The Ariane 5 Flight 501 inquiry report states the cause as a software
exception in a routine that converted a 64-bit floating-point number
to a 16-bit signed integer \cite{acc:ariane501-ifr}. Argue, in 200
to 300 words, whether this is a measurement failure, a software
failure, or a verification failure. State which framing produces the
most useful corrective action.

\paragraph{Solution.}
We argue each framing in turn, evaluate the corrective action it
produces, and identify the framing whose corrective covers the
broadest part of the failure chain.

\textit{Step 1, frame as measurement failure.} The Inertial Reference
System (SRI) was measuring horizontal velocity, an internal quantity
used in the alignment routine that had been carried over unchanged
from Ariane 4. On Ariane 5, the booster's higher trajectory caused
horizontal velocity to exceed the maximum value representable as a
16-bit signed integer at $T+37 \,\si{\second}$. The measured quantity
left the range its representation could hold; the type-conversion
generated an operand-error exception. Read as measurement, the
failure is a range-mismatch between the measurand and its
representation. The engineering discipline of stating the maximum and
minimum range of every measured internal quantity, and verifying that
the representation is wide enough to hold the range under the
intended flight envelope, would have surfaced the mismatch.

\textit{Step 2, frame as software failure.} The conversion from a
64-bit floating-point value to a 16-bit signed integer was
unprotected by a range check or exception handler. The handler that
fired at the overflow followed the system's default behaviour of
shutting down the SRI; the backup SRI was running the same software
and shut down for the same reason at the same flight time; the
launcher's guidance system received nonsense data and commanded full
nozzle deflection; the aerodynamic loads exceeded structural limits
and the launcher self-destructed. Read as software, the failure is
an unguarded conversion in a critical path with dissimilar-software
redundancy absent. Defensive software practices (range checks,
exception handlers, type-conversion guards, dissimilar redundancy)
would have prevented the cascade from the operand-error to the
loss of vehicle.

\textit{Step 3, frame as verification failure.} The SRI software was
qualified for the Ariane 4 flight envelope, in which horizontal
velocity remained within representable range. Ariane 5 had a
different trajectory profile, with higher horizontal velocity
earlier in the flight. The software was reused on Ariane 5 without
re-qualification against the new envelope; the alignment routine
was running redundantly post-liftoff for legacy reasons specific to
Ariane 4's launch-pad alignment procedure, and was not needed on
Ariane 5. Read as verification, the failure is the absence of a
qualification-envelope match between the inherited software and the
new vehicle.

\textit{Step 4, evaluate the corrective for each framing.} The
measurement framing produces the corrective ``add range checks to
internal-quantity measurements.'' This is a narrow fix; it addresses
the SRI's representation but leaves the broader question of when
inherited subsystems should be re-qualified unaddressed. The software
framing produces the corrective ``add defensive guards and dissimilar
redundancy.'' This is broader; it addresses the cascade but does not
catch the fact that the alignment routine should not have been
running on Ariane 5 at all. The verification framing produces the
corrective ``re-qualify every inherited subsystem against the new
vehicle envelope before integration.'' This is the broadest; it
catches the unnecessary routine, the range mismatch, and the
software gap as a single point of audit.

\textit{Step 5, identify the broadest-coverage framing.} The
verification framing covers the most of the failure chain because the
proximate engineering decision that produced the failure was the
decision to reuse the qualified Ariane 4 SRI software on Ariane 5
without re-qualification against the new flight envelope. Both subordinate
correctives (range checks and defensive software) are sub-cases of
the verification corrective; an envelope-driven re-qualification of
the SRI module on Ariane 5 would have either added the range checks
as part of the re-qualification, removed the alignment routine as
unnecessary, or both.

\textit{Step 6, state the broader pattern.} Inherited subsystems
carry their qualification envelopes with them, and the qualification
envelope is part of the subsystem's specification. When a subsystem
is reused on a new vehicle, the envelope must be re-evaluated against
the new vehicle's profile; the absence of an envelope-match check is
the failure mode that this and several subsequent inherited-software
accidents share. The corrective is procedural and verifiable; it
makes the envelope visible at the integration interface, which is
where the inheritance decision is actually made.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
The reader is presented with a beam designed to carry $10 \,\si{\kilo
\newton}$. The drawing is dimensioned in millimetres; the
specification is in newtons; the supplier's catalogue gives
dimensions in inches and allowable loads in lbf. Argue, in one
paragraph, what you do.

\paragraph{Discussion.} The working engineer converts the supplier
catalogue once, in writing, at the procurement interface, and
proceeds with the project in SI. The conversion is performed by a
second engineer or by an automated tool and the result is recorded
in the bill of materials with both the source value and the
converted value, so the audit trail survives. The temptation to
``work in inches because the supplier does'' is the failure mode
the chapter has described: every later calculation that touches
the beam's dimensions or load capacity would otherwise carry the
unit-conversion risk. The procurement interface is the right
location for the conversion, because it is the only point at which
the supplier's catalogue intersects the engineering work.
\end{exercise}

\begin{exercise}
A senior colleague tells the reader that ``the measurement is
obvious; you do not need to write down the conditions.'' Write a
one-paragraph reply that takes the colleague seriously, agrees where
agreement is honest, and disagrees where disagreement is honest.

\paragraph{Discussion.} A credible reply concedes that some
measurements have stable conditions that genuinely do not require
recording for the immediate purpose (a routine calibration check
in a temperature-controlled lab under known protocol). It
disagrees on the broader principle: the recording serves the next
engineer, not the current one, and the current engineer is rarely
the one who has to defend the measurement after a failure. The
recording is also a defence against the engineer's own future
self, who will not remember the conditions in six months. The
discipline is to record by default and document the exceptions
deliberately, rather than skip by default and document only when
remembered. The senior colleague's point is valid for routine work
in well-characterised conditions; the chapter's discipline is
correct for the cases the routine eventually generalises to.
\end{exercise}

\begin{exercise}
Identify a measurement in your immediate environment that you have
been making for years without realising it (e.g., judging when the
kettle is hot enough to brew tea, judging when the car is too far
from the curb, judging when an email is too long). State the implicit
quantity, the implicit instrument, and the implicit uncertainty.
Comment on whether this measurement could be improved by being made
explicit.

\paragraph{Discussion.} The exercise is open by design; the
reader's answer is the deliverable. A representative response: the
implicit quantity is the temperature of the kettle water; the
implicit instrument is the sound of the rolling boil; the implicit
uncertainty is the difference between the boiling-noise threshold
and the actual $100 \,\degC$, perhaps $\pm 2 \,\degC$. The
measurement could be made explicit with a thermometer for tea-
specific temperatures (green tea at $\sim 75 \,\degC$ benefits from
explicit measurement; black tea at $100 \,\degC$ does not). The
broader pattern: the craftsman's knowledge of section 1.2 is often
precise but unstated; making it explicit converts it into the
engineer's knowledge that can be audited, transferred, and
improved. The conversion is not always worth the effort, but it is
worth recognising the conversion exists.
\end{exercise}

\begin{exercise}
Argue, in 200 to 300 words, whether the seven-component measurement
habit described in section 1.5 is too much for a working engineer
to maintain at the pace of real engineering. Take a position. Defend
it.

\paragraph{Discussion.} A credible argument names the actual cost
(the seven-component recording takes $\sim 30 \,\text{seconds}$
per measurement at a calibration log; in a production environment
this can become $\sim 10$ minutes per shift across all
measurements) and weighs it against the benefit (audit-defensibility,
diagnostic capability after a failure, traceability for regulatory
purposes). A position that the discipline is too much is defensible
in production environments where the regulatory and audit
requirements do not justify the cost (e.g., routine hobbyist
measurements); it is indefensible in production engineering where
failure consequences are non-trivial. A position that the discipline
is appropriate is defensible in any setting where the measurement
could be the subject of a future inquiry; the cost of recording is
small compared to the cost of an undefended failure. The chapter's
position is that the discipline is appropriate by default with
exceptions documented deliberately. The reader's argument is
credible if it engages with the cost-benefit at the level of
specific engineering settings rather than at the level of
principle.
\end{exercise}
