\chapter{Forensic engineering and accident investigation}
\label{vol10:ch12}

\epigraph{The accident has happened; the evidence is degrading; the work is to extract from it the truth that will prevent the next one.}{}

\chapmeta{Half-life: durable. Archetypes: ethics, failure. Exercise target: 20. Project track: hybrid.}

\noindent
Forensic engineering is the discipline of reconstructing engineering failures: determining what failed, how it failed, in what sequence, under what loads, with what contributory factors. The investigation is constrained by the evidence that survived the failure, by the access the investigators have to the wreckage and the data, by the legal and commercial pressures of the parties affected, and by the time available before the evidence degrades or is destroyed. The product is the report that the safety community, the regulators, the courts, the insurers, and the next generation of designers will use to understand the failure. This chapter is about the discipline: how the work is done, what it produces, what protects it from error, what makes it sometimes wrong.

\section{The role of the forensic engineer}\pathtag{core}

The forensic engineer's role differs from the design engineer's in several axes. The design engineer works forward (from intended function to constructed system); the forensic engineer works backward (from observed failure to causal sequence). The design engineer has access to the engineering team's institutional knowledge; the forensic engineer often must reconstruct the design intent from documents and from the system itself. The design engineer's product is a system; the forensic engineer's product is a report. The design engineer's accountability is to the user of the system; the forensic engineer's accountability is to the public's interest in understanding the failure, which may be served by reports the parties affected do not want published.

The forensic engineer's clients are typically several:

\textbf{Safety investigation authorities} (NTSB, AAIB, BEA, JTSB, ATSB, KNKT, CSB, the various national investigative bodies for nuclear, marine, and other incidents). These bodies have statutory authority to conduct safety investigations and to issue reports that the public can rely on. Their charters typically separate them from regulators (who have enforcement authority) so that the safety investigation can proceed without being affected by, or affecting, enforcement decisions.

\textbf{Regulators} (FAA, FDA, EPA, OSHA, the various national equivalents). These bodies investigate to determine whether regulations were violated, whether enforcement action is warranted, and whether the regulatory framework needs revision. The regulator's investigation may be in parallel with a safety investigation; the relationships between regulator and investigation authority are typically structured so that the safety report's conclusions are not held back by the regulator's enforcement processes.

\textbf{Insurers and reinsurers.} The investigation that establishes the cause of a loss informs the claims process, the subrogation analysis (recovery from other parties whose contribution to the failure can be established), and the underwriting of future risks. Insurance-driven investigations are private; their reports are usually not published.

\textbf{Litigants.} Plaintiffs in product-liability suits, defendants in criminal proceedings against engineering organisations, and parties to commercial disputes all retain forensic engineers as experts. The litigation context shapes the investigation (focused on the parties' specific claims), the report (often privileged communication; not public unless the case proceeds to trial), and the engineer's testimony (which must withstand cross-examination on methodology and conclusions).

\textbf{The engineering profession itself.} The published safety reports and the academic case-study literature are the corpus that the next generation of engineers learns from. The forensic engineer's responsibility to the profession is to contribute to that corpus accurately and thoroughly.

The engineer's professional obligations across these clients are aligned in principle (find the truth; report the truth) but may be in tension in practice. The investigation conducted for a litigant may not be permitted to report findings adverse to the litigant's case; the investigation conducted for an insurer may not produce a public report. The professional codes of ethics (NSPE, IEEE, ASCE, ASME, the international equivalents) provide the structure within which the tensions are managed; the engineer who finds them irreconcilable in a specific case may need to withdraw.

\section{Evidence preservation}\pathtag{core}

The accident scene begins degrading the moment the accident ends. Wreckage is moved, weather acts on it, parties with interests in the outcome interact with it, and the time-critical evidence (fluid levels, fluid samples, surface conditions, temporary configurations, transient electronic state) becomes irretrievable. The first investigative action is preservation; the quality of the eventual report depends on what was preserved in the hours after the event.

The preservation actions, in approximate order of urgency:

\textbf{Scene security.} The accident scene is secured against entry by unauthorised parties: cordoned off, photographed in its as-found state, with access logged. Aviation accident scenes are typically secured by local law enforcement until the investigation authority arrives; industrial accident scenes by the operator's emergency response; structural collapse scenes by the police and fire services. The security is contested in practice: parties with interests in the outcome (the equipment manufacturer; the operator; the insurer; the parties' lawyers) may attempt to access the scene before the investigation establishes control.

\textbf{Photographic and video documentation.} The scene is documented before anything is moved: wide-angle context shots; medium-range component shots; close-up detail of significant features; orthogonal views; scale references; standardised lighting. Modern practice extends to laser scanning (3D point clouds of the entire scene; structure-from-motion photogrammetry from drone-collected imagery) that produces a digital reconstruction that can be revisited indefinitely. The documentation is the only evidence that survives wreckage removal in its original spatial relationships.

\textbf{Recorder recovery.} The flight data recorder and cockpit voice recorder in aviation; the train event recorder; the engine control module in automotive; the SCADA historian in process industries; the application logs in software systems. These devices and the data they contain may be the most complete record of the accident's sequence; their physical recovery (often from severe damage; sometimes from underwater) is a specialised activity. The data is then transferred to read-out facilities that can recover the data even from damaged storage. The AF447 flight data and cockpit voice recorders, recovered from the Atlantic Ocean floor in 2011 after a multi-year search, were the evidence that enabled the investigation's substantive findings \cite{acc:bea-af447}.

\textbf{Wreckage cataloguing.} Each significant piece of wreckage is identified, tagged with a unique identifier, located in space (typically on a scaled scene plan or in a 3D point cloud), photographed, and (where appropriate) removed to a controlled facility for further examination. The catalogue is the basis for subsequent reconstruction; large debris fields (such as Columbia's, spread across two states) require systematic gridded search and recovery. The Columbia debris reconstruction at Kennedy Space Center, in which the recovered orbiter debris was laid out on a floor in approximate as-flown position, is the canonical example of post-accident wreckage reconstruction at scale.

\textbf{Sample preservation.} Materials samples (metallurgical specimens; chemical samples; biological samples; soil samples) are taken under chain-of-custody procedures so that their provenance can be established in subsequent analysis and (if needed) in legal proceedings. The chain-of-custody record documents who handled the sample, when, where it was stored, and what was done to it.

\textbf{Witness statements.} People who saw the accident, who were involved in operating the system, who were responsible for its maintenance, are interviewed promptly. Memory degrades and contamination by post-event information accumulates; the first statements are typically the most reliable. The interview protocol is itself a discipline; trained investigators use open-ended questioning, avoid leading questions, document the interview in writing, and where possible record it.

\textbf{Electronic record preservation.} Server logs, network captures, configuration management records, email archives, design-document version histories. These records are the evidence of the engineering and operational decisions that led to the failure. Modern legal hold procedures (in which the parties affected are notified of an obligation to preserve potentially relevant electronic records) are the institutional mechanism for protecting against post-event spoliation; the obligation is enforceable in court.

The failure modes in evidence preservation are several and consequential. Scene contamination (parties with interests in the outcome handle wreckage before investigation control); recorder damage (the recorder design did not survive the failure conditions); sample mis-identification (chain-of-custody errors); witness contamination (witnesses talk to each other and to media before being interviewed); record destruction (the deletion of relevant electronic records, sometimes automatic per data-retention policy, sometimes deliberate). The investigative discipline includes monitoring for these failure modes and adjusting subsequent analysis to account for them.

\section{Reconstruction methods}\pathtag{core}

Once evidence is preserved, the analytic work begins. The methods are partly common across domains and partly domain-specific. The common-across-domains methods:

\textbf{Time-sequence reconstruction.} The events are ordered in time using recorder data, witness statements, physical evidence (impact marks; debris fields; environmental disturbances), and the natural ordering imposed by physics (an event that produces a particular effect must precede that effect). The reconstruction often resolves down to fractions of seconds; aviation accident sequences are typically reconstructed to one-second resolution or better.

\textbf{Energy accounting.} The energy in the system before the accident and the energy released during the accident are accounted for. Conservation laws are absolute; an unexplained energy gap indicates that the reconstruction is incomplete. The energy accounting was central to the Columbia investigation \cite{acc:caib-columbia}: the kinetic and thermal energy budgets ruled out and ruled in possible damage scenarios for the wing leading edge.

\textbf{Failure-tree construction.} The accident's terminal state is decomposed into the immediate causes that could have produced it, each of those into their immediate causes, recursively until causes are reached that are either independently verifiable from evidence or that are accepted as the root of the analysis. The tree's branches that are inconsistent with evidence are pruned; the branches that are consistent are retained as candidate scenarios. The process resembles the design-time fault-tree analysis (Chapter 1) but is conducted in reverse against actual evidence rather than against hypothetical failure modes.

\textbf{Hypothesis testing.} Each candidate scenario predicts specific observable evidence; the investigation searches for that evidence; the scenarios consistent with the observed evidence are retained, the inconsistent are rejected. The discipline is Popperian: a scenario that explains everything observed is interesting; a scenario that predicts an observation that the investigation can search for and find or not find is more interesting; a scenario that survives multiple such tests is the working hypothesis. The investigation does not stop at the first plausible scenario; it tests against all available evidence and against alternative scenarios.

\textbf{Simulation.} The candidate scenarios are tested in simulation: computational fluid dynamics for aerodynamic events; finite-element analysis for structural events; full-vehicle dynamics for automotive events; full-system simulation for control-system events. The simulations are validated against the recorded data and against the physical evidence; a simulation that matches the data supports the scenario the simulation embodies. The simulation is also a way to test the scenario's sensitivity: how would the outcome have changed under different conditions; would the scenario have produced the observed evidence under reasonable parameter ranges.

\textbf{Re-creation.} Some investigations include physical re-creation: a test article subjected to the conditions the scenario predicts, instrumented to observe the resulting failure. The Comet re-creation in the post-Comet investigation (Chapter 2) tested a complete fuselage in a water tank, cycling pressurisation to reproduce the in-service fatigue history \cite{acc:cohen-comet-report-1955,text:withey-fatigue-comet}. The re-creation produced the same fatigue cracks at the same locations as the in-service failures, establishing the mechanism beyond reasonable dispute.

\textbf{Cross-domain validation.} The reconstruction's conclusions are reviewed against the engineering literature, against the experience of similar systems, and against the judgment of subject-matter experts not directly involved in the investigation. The review is the discipline's defence against the in-investigation team's specific blind spots.

\section{Metallurgical, electrical, structural forensics}\pathtag{standard}

The domain-specific reconstruction methods are extensive; the following are representative.

\textbf{Metallurgical forensics.} The examination of failed metallic components establishes the failure mode (ductile overload; brittle fracture; fatigue; corrosion; stress-corrosion cracking; creep; hydrogen embrittlement; thermal shock). The diagnostic evidence is visible on the fracture surface: ductile failure shows microvoid coalescence under scanning electron microscope; brittle failure shows cleavage facets; fatigue shows beach marks and striations; corrosion shows oxide and pit morphology; stress-corrosion cracking shows intergranular branched fracture. The metallurgist also examines the material composition (against the certified composition), the microstructure (grain size, phase distribution, heat-treatment indicators), and the surface treatment. Volume II Chapters 1-5 established the materials methods; this chapter's use is reverse: from observed failure to the mechanism and the contributing conditions. Chapter 2 of this volume (fatigue) and Chapter 3 (corrosion) are the substrate for the metallurgical forensic work in those failure modes.

\textbf{Electrical and electronic forensics.} The examination of electrical components establishes the failure mode (overcurrent; overvoltage; insulation breakdown; thermal damage; mechanical damage). Burnt or carbonised components, melted conductors, arc tracking on insulators, and signature damage patterns on semiconductor dies (under scanning electron microscope or focused ion beam cross-section) provide diagnostic evidence. Recorded data from instrumentation, where available, provides the time history of voltages and currents leading to the failure. The electronics forensic discipline has developed substantially in the era of pervasive embedded systems; the recovery of state from damaged or destroyed devices (flash memory recovery from physically damaged storage; firmware extraction from boot ROM; oscilloscope captures of pre-failure behaviour from instrumented test setups) is now a routine part of consumer-electronics failure analysis as well as accident investigation.

\textbf{Structural forensics.} The examination of failed structures establishes the load path that produced the failure, the magnitude of the loads, and the conditions (material strength, geometry, support conditions) that determined the strength. The diagnostic evidence is the spatial pattern of damage: where the crack initiated, how it propagated, what surrounding structure failed in what sequence. The Hyatt Regency walkway collapse \cite{acc:nbs-hyatt-1982}, the Tacoma Narrows Bridge failure \cite{acc:tacoma-narrows-1941}, and the Millennium Bridge instability \cite{acc:dallard-millennium-2001} are canonical structural-forensic cases; each is documented in this book's earlier volumes.

\textbf{Chemical and process forensics.} The examination of process equipment after a chemical accident establishes the substances involved, the conditions under which the accident proceeded, the path the released material took, and the contribution of equipment failures to the release. The Bhopal investigation (Chapter 10, \cite{acc:icmr-bhopal-2004}) and the BP Texas City investigation \cite{acc:csb-bp-texas-city-2007} are extensive examples of chemical and process forensic work; the CSB's Imperial Sugar investigation \cite{acc:csb-imperial-sugar-2009} is the canonical recent case for combustible-dust forensic methodology.

\textbf{Marine and offshore forensics.} The examination of failed marine and offshore structures, often conducted underwater or after recovery to surface, establishes the load history (wave loading, structural impact, corrosion), the structural failure sequence, and the contribution of operational and procedural factors. The investigation environment is hostile (depth, currents, biofouling, time pressure to recover) and the techniques are correspondingly specialised (ROV-based survey; saturation diving; recovered-debris analysis at surface facilities).

\textbf{Geotechnical forensics.} The examination of slope failures, foundation failures, and earthworks failures establishes the soil and rock conditions, the loads, and the failure surface. Geotechnical investigation is often time-critical (the failure may be still evolving), spatially extensive, and dependent on instrumentation installed before or rapidly after the event.

\section{Software forensics}\pathtag{core}

Software failures present forensic challenges different from physical-system failures. The evidence is electronic and may be ephemeral (a system that has crashed may have lost the state that would have shown the failure mode; a system that has been rebooted has lost the in-memory evidence; a system whose logs have rolled over has lost the record of pre-failure activity). The reconstruction depends on what was logged, what was captured by instrumentation, and what can be recovered from the affected systems.

The methods:

\textbf{Log analysis.} Application logs, system logs, network logs, security logs, and (where they exist) audit logs are the primary record of system behaviour. The forensic work assembles the logs in time order, correlates events across multiple systems' logs (which may differ in clock and in time-zone), and reconstructs the sequence of operations that preceded the failure. Modern observability tooling (centralised log aggregation; distributed tracing; metrics history) has substantially improved the available evidence, but log retention policies, log levels (the volume of detail logged), and log integrity (protection against post-incident tampering) remain design choices that constrain forensic capability.

\textbf{Memory and disk image analysis.} A failed system's memory and disk state, captured at the time of failure or as soon thereafter as possible, provides the in-system evidence of the failure. Memory captures may include the call stacks of failed threads, the state of in-flight requests, the contents of caches and queues, and (where the failure mode involves memory corruption) the corrupted data itself. The Knight Capital 2012 incident's investigation included reconstruction of the dormant code that was unintentionally activated; the activation pattern and the resulting trading behaviour were reconstructed from the affected systems' state at the time of the incident.

\textbf{Network capture analysis.} Packet captures of the network traffic into and out of the affected systems provide the external evidence of the failure: the requests that triggered the failure, the responses that propagated it, the protocol-level patterns. Network forensic capability requires that the captures exist; many production environments do not capture sufficient detail for forensic reconstruction by default.

\textbf{Code archaeology.} The version of the code that was running at the time of the failure is reconstructed from the version control system, from the deployment records, and (where necessary) from the binaries on the affected systems. The reconstruction may require resolving against ambiguous deployment histories, against partially-applied configuration changes, and against the difference between source code and what actually ran.

\textbf{Reproduction.} The failure is reproduced in a controlled environment by recreating the input sequence, the state, and the system configuration that triggered it. Reproduction is the gold standard for understanding software failures (it confirms the mechanism unambiguously), but is often not feasible for complex production failures whose triggering state cannot be fully reconstructed. The Ariane 501 floating-point overflow failure \cite{acc:ariane501-ifr} was reproduced in subsequent simulation; the Therac-25 race condition was reproduced after the mechanism had been identified.

\textbf{Static analysis of the involved code.} The code is examined for the failure mode: the race condition that produced the deadlock; the unchecked numeric conversion that produced the overflow; the unbounded recursion that exhausted the stack; the input pattern that produced the buffer overrun. Modern static-analysis tools (Coverity, Klocwork, the various sanitisers, formal verification systems) accelerate this work; the result of the analysis is the identification of the specific code defect and the conditions under which it manifests.

The post-Therac-25 software-forensic literature has matured into a discipline; Leveson's \emph{Engineering a Safer World} (2011) and \emph{Safeware} (1995) document the methodology, with case studies that have become the field's standard references \cite{text:leveson2011,text:leveson-safeware}. The lessons aggregate to architectural recommendations (logging discipline; observability; reproducibility of production state in test environments) that good engineering practice has incorporated and that subsequent failures continue to surface.

\section{Reporting standards: NTSB, AAIB, BEA}\pathtag{standard}

The reports produced by major safety investigation authorities are themselves engineering artefacts with developed conventions, standard sections, and structural protections against bias.

The NTSB (US National Transportation Safety Board) reports use a standard structure: factual information (the established facts; reviewed by the involved parties for accuracy without their being able to change them); analysis (the NTSB's reasoning about the factual information; not subject to party review); conclusions (the findings; the probable cause; contributing factors); recommendations (the actions the NTSB recommends, addressed to specific organisations). The probable-cause statement is brief and unambiguous; the analysis section is detailed and shows the reasoning. The party process gives manufacturers, operators, and other involved organisations early access to factual information and the opportunity to contribute technical expertise, without giving them control over the analysis or conclusions; this is a structural protection against the regulatory-capture-style dynamics in investigation.

The AAIB (UK Air Accidents Investigation Branch) and the BEA (French Bureau d'Enquêtes et d'Analyses) follow similar structures, with national variations. International coordination of aviation accident investigation is provided by ICAO Annex 13, which establishes the standards for state-of-occurrence, state-of-registry, state-of-operator, state-of-design, and state-of-manufacture participation; the standards align the reports across jurisdictions sufficiently that the global aviation safety community can use them comparatively.

The CSB (US Chemical Safety and Hazard Investigation Board) follows a similar structure adapted for process-industry incidents; CSB reports often include detailed animations and videos that communicate the technical mechanism to non-specialist audiences. The CSB's communication discipline has been influential; many other investigation authorities have adopted similar visual communication standards.

The NRC (US Nuclear Regulatory Commission) and the IAEA (International Atomic Energy Agency) produce reports following the conventions of nuclear-safety regulatory practice; the INSAG series (International Nuclear Safety Advisory Group) reports on major events are the canonical international references \cite{acc:iaea-insag1-1986,acc:iaea-insag7-1992}.

The common discipline across these reporting standards:

\textbf{Separation of factual from analytical.} Facts are presented before analysis; the reader can evaluate the analysis against the facts; subsequent analyses can be conducted using the same facts.

\textbf{Sourcing of claims.} Specific claims (numerical values, dates, sequences) are sourced to specific evidence (data recorder values, witness statements, physical evidence). The sourcing allows the claim to be re-evaluated against the same evidence by later analysts.

\textbf{Recommendation tracking.} Recommendations are addressed to specific organisations, with specific actions requested, and the investigation authority maintains a public record of the recommended actions' status (open, closed, accepted as alternative, not accepted). The NTSB's "Most Wanted List" is the canonical example of recommendation-tracking discipline.

\textbf{Independence and transparency.} The investigation authority is institutionally independent from the regulators it makes recommendations to; the reports are public; the investigation process is described in the report; the dissents (where they exist) are noted.

\section{Legal and ethical pressures on forensic engineers}\pathtag{standard}

Forensic engineering is conducted in a contested space. The parties affected by the investigation's outcome (manufacturers facing liability; operators facing regulatory consequences; insurers facing payment obligations; governments facing political consequences) have strong incentives to influence the investigation. The engineer's professional obligations resist that influence.

The pressures that recur:

\textbf{Client pressure on conclusions.} The engineer's client wants conclusions that support the client's interest; the engineer's professional obligation is to the truth that the evidence supports. The conflict is sometimes resolved by the engineer's withdrawal from the engagement (a routine professional move when the engineer cannot ethically reach the conclusions the client wants); sometimes by the engineer's report stating the conclusions the evidence supports regardless of the client's preference (with the consequences for the client managed by the client's lawyers); sometimes by negotiated reports whose silences and emphases reflect the compromise without containing falsehoods.

\textbf{Pressure to extend findings beyond evidence.} The engineer's report is sometimes wanted to support claims the evidence does not actually support: the conclusion that some party is responsible for some outcome that the evidence shows occurred but does not clearly attribute. The engineer's obligation is to state what the evidence supports and to identify what it does not; the boundary is what separates engineering analysis from advocacy.

\textbf{Pressure to suppress findings adverse to the client.} The engineer's report is sometimes wanted to omit findings that would embarrass the client or expose the client to consequences. The engineer's professional obligation is to report what the evidence shows; the contractual privilege of the client-engineer relationship is sometimes invoked to keep the report confidential, but the privilege does not authorise the engineer to falsify the report.

\textbf{Testimony under cross-examination.} The engineer testifying in legal proceedings is subjected to adversarial questioning designed to undermine the engineer's analysis or credibility. The discipline is preparation: the engineer must know the evidence in detail, the methodology in detail, the alternative scenarios in detail, and must be prepared to acknowledge the limits of the analysis honestly. Engineers who overclaim under cross-examination undermine the report's credibility; engineers who acknowledge the limits maintain it.

\textbf{Conflict of interest.} The engineer who has consulted for the manufacturer, the operator, or other affected parties on related matters may have a conflict that disqualifies them from forensic engagement on the current matter. Disclosure of prior relationships is a routine professional obligation; the threshold for disqualification depends on the specifics.

\textbf{Public attention.} High-profile accidents attract media attention; the investigation team faces pressure to issue preliminary conclusions before the analysis supports them, to comment on the case in ways that prejudice subsequent legal proceedings, or to release information that the investigation authority has reasons to withhold. The discipline of investigation authorities is to channel public communication through formal releases, to maintain professional silence on the specifics until the report is ready, and to resist the pressure for premature conclusions.

The codes of ethics of the engineering professional bodies address these pressures explicitly. NSPE's code (the National Society of Professional Engineers, US) includes specific provisions on objectivity in reports, on conflict of interest, and on professional integrity in expert testimony. The IEEE, ASCE, ASME codes have analogous provisions. The professional registration that licenses Professional Engineers in many jurisdictions includes enforceable obligations under these codes; engineers who violate them face professional sanction (suspension or revocation of registration) as well as the loss of professional standing.

\section{Failure: when forensic engineering reached the wrong conclusion}\pathtag{core}

Forensic engineering is itself fallible. The cases in which the discipline has reached the wrong conclusion are instructive both as cautionary examples and as illustrations of the structural conditions that protect against error.

The IAEA's initial INSAG-1 report on Chernobyl \cite{acc:iaea-insag1-1986} attributed the accident substantially to operator violations of operating procedures; the INSAG-7 update in 1992 \cite{acc:iaea-insag7-1992} substantially revised the assessment, identifying the reactor's design properties as primary causal factors. The shift is the canonical example of a major investigation reaching the wrong conclusion the first time and revising it; the conditions that allowed the revision included the post-Soviet release of technical information about the RBMK design, the broader international participation in the revised analysis, and the institutional independence of the IAEA from the Soviet nuclear establishment. The lesson is partly the importance of evidence access (the initial report had limited access to design information that the revised report could draw on) and partly the importance of independent international review (the revision was conducted by an international advisory group rather than by the Soviet nuclear authorities alone).

The 1985-1989 investigation of the Aloha Airlines 243 incident (the Boeing 737 fuselage failure of 1988 \cite{acc:ntsb-aar94-04}) initially focused on metal fatigue as the primary mechanism, with the disbonded fuselage skin as a contributory factor. The investigation's subsequent work, drawing on post-incident inspections of the fleet, identified multi-site fatigue cracking driven by both fatigue cycling and disbond of the cold-bonded fuselage joints as the joint mechanism; the relative weight of the two contributors and the manufacturer's role in the disbond problem evolved through the investigation. The case is not one of "wrong conclusion" reached and reversed, but of the depth of analysis evolving as the investigation proceeded; the lesson is that early conclusions in major investigations are provisional and the discipline benefits from publishing the full analysis with its evolution visible.

The Hubble Space Telescope spherical-aberration investigation \cite{acc:nasa-hubble-optical-systems-1990} reached the correct technical conclusion (the primary mirror's null corrector had been mis-assembled, producing a polishing error) and the correct attribution (the manufacturer's failure to follow established quality-control procedures), but the institutional response (NASA's procurement and oversight changes) was driven in part by political pressure rather than by the technical findings alone. The case illustrates the distinction between the forensic engineering report (which should be technical and accurate) and the institutional response (which is subject to political pressures the report cannot itself constrain).

The structural protections against wrong conclusions in forensic engineering:

\textbf{Independent technical review.} The investigation's findings are subject to review by technical experts not directly involved in the investigation. The review may be internal to the investigation authority (a peer review before publication), external (academic engineering commentary; industry technical commentary), or institutional (the IAEA's revision of INSAG-1 to INSAG-7).

\textbf{Reproducibility of analysis.} The investigation's evidence is preserved and is accessible (subject to legal and proprietary constraints) to subsequent analysts. The reproducibility allows the analysis to be challenged on its substance rather than only on its conclusions.

\textbf{Recommendation tracking.} The investigation's recommendations are tracked over time; the tracking surfaces the conditions in which the recommendations were inadequate or not implemented.

\textbf{Cross-investigation comparison.} Similar accidents are compared; recurring patterns surface that single-investigation analysis may miss. The CAIB's identification of Challenger-like organisational patterns at Columbia \cite{acc:caib-columbia} is the canonical example.

\textbf{Standards revision.} The reporting standards themselves are revised in response to forensic-engineering failures; ICAO Annex 13's evolution reflects accumulated lessons from major aviation investigations.

\begin{principle}[Honest report]
The forensic engineer's responsibility is the report that the evidence supports. Reports that exceed the evidence (overclaiming attribution or mechanism), or that under-report it (omitting findings adverse to the client), corrode the corpus that the next generation of engineers learns from. The discipline of honest reporting under contested conditions is what makes forensic engineering useful to the broader engineering profession.
\end{principle}

\begin{archetype}[failure]
Forensic engineering is the archetype's after-action study. The failure has occurred; the discipline is to extract from it what the next iteration of the system must change. The work is honest engineering analysis under the pressures of contested interests; its quality determines whether the field learns from the accident.
\end{archetype}

\begin{project}[Forensic investigation of a small failure]
Select one small failure to investigate. Options: a broken household object (a chair leg that snapped; a cracked plate; a failed power adapter; a torn bicycle tube; a burned-out light bulb); a software crash on the reader's own machine (a kernel panic; an application that exited unexpectedly; a build that failed in an unfamiliar way); a small mechanical failure observed in the reader's life (a door hinge that came off; a zipper that failed; a fastener that backed out). Conduct a forensic investigation following the discipline of this chapter: preserve the evidence; document the as-found state with photographs and notes; reconstruct the sequence of events; identify the failure mechanism; identify the contributing factors; identify what could be changed to prevent recurrence. Produce a five-page report following the structure of an NTSB-style report: factual information; analysis; conclusions including probable cause; recommendations.

\textbf{Hazard.} The exercise is deliberately scaled to a low-stakes failure so the reader can practice the discipline without the pressures and constraints that real forensic engagements carry. The report's professional register should match a published investigation report's; the analytical depth should be substantial despite the failure's modest scale.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A wreckage debris field from an aircraft accident extends $5\,\text{km}$ along the flight path and $200\,\text{m}$ across. Estimate the number of pieces in a search grid of $20\,\text{m} \times 20\,\text{m}$ cells. If each cell takes 1 hour to search and document, compute the total search effort in person-hours and the calendar time required for a team of 50 searchers.
\end{exercise}

\begin{exercise}
A flight data recorder samples airspeed at 4 Hz. An aircraft loses airspeed indication, then recovers it 6 seconds later. Compute the number of samples missing. Discuss how the time resolution of the FDR constrains the reconstruction of fast transients.
\end{exercise}

\begin{exercise}
A fatigue crack on a failed aircraft component shows striation spacing of $1\,\mu\text{m}$ in one region and $10\,\mu\text{m}$ in another. Treating each striation as one load cycle, compute the cycle count for a $5\,\text{mm}$ crack region in each spacing regime. Discuss what the spacing variation indicates about load history.
\end{exercise}

\begin{exercise}
A software log captures events at 100 events per second from each of 50 servers. Compute the storage required to retain logs at full detail for 30 days. Discuss the operational trade-offs between log retention duration and storage cost.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the expected reconstruction accuracy of a stereoscopic photogrammetric system as a function of the baseline length between cameras, the focal length, and the object distance. Identify the geometric regime in which sub-millimetre accuracy at 10 m range is feasible.
\end{exercise}

\begin{exercise}
For a failure-tree reconstruction with $N$ candidate scenarios, each predicting a different set of evidence, derive the expected number of evidence tests required to identify the correct scenario. Identify the conditions under which more tests are required than the available evidence supports.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the duration of the Columbia debris reconstruction effort at Kennedy Space Center (the floor reconstruction of the recovered orbiter debris) \cite{acc:caib-columbia}. Identify the labour and infrastructure required.
\end{exercise}

\begin{exercise}
Estimate the cost of a major-incident forensic investigation by a national authority (NTSB, AAIB, BEA). Compare to the cost of the institutional changes the investigation typically produces.
\end{exercise}

\begin{exercise}
Estimate the time required for a forensic software investigation of a production-system outage that involves three teams, 50 servers, and three weeks of log data. Identify the dominant time-consumers.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Implement a simulation of an evidence-degradation process: a forensic scene has $N$ pieces of evidence; each piece degrades to unreadable at independent random times drawn from an exponential distribution with mean $\tau$; the investigation team can preserve $k$ pieces per hour. Plot the fraction of evidence preserved as a function of team arrival time at the scene.
\end{exercise}

\begin{exercise}
Simulate a recorder-recovery scenario: a flight data recorder is at depth $d$ in the ocean; ROV search proceeds at $0.1\,\text{km}^2$/day; the search area is the uncertainty ellipse from the last-known-position calculation, with axes proportional to $\sqrt{t}$ where $t$ is the time since loss-of-contact. Plot expected recovery time as a function of search-start delay.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design an evidence-preservation protocol for a process-industry incident response team. Specify the actions in the first 30 minutes, the first 4 hours, and the first 24 hours.
\end{exercise}

\begin{exercise}
Design a logging architecture for a distributed system that maximises forensic capability while keeping storage cost bounded. Specify the log levels, the retention durations, the integrity protection, and the access controls.
\end{exercise}

\begin{exercise}
Design a forensic-investigation report template for a hypothetical organisation. Include the sections that the NTSB-style structure recommends, the source-tracking discipline, and the recommendation-tracking field.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
A forensic engineer's client (an equipment manufacturer) asks the engineer to omit a finding (a design defect contributed to the failure) from the engineer's report. Discuss the engineer's options under the NSPE code of ethics. Identify the conditions under which withdrawal is required versus reporting the finding regardless of the client's preference.
\end{exercise}

\begin{exercise}
The IAEA's INSAG-1 report on Chernobyl \cite{acc:iaea-insag1-1986} reached conclusions that were substantially revised by INSAG-7 in 1992 \cite{acc:iaea-insag7-1992}. Discuss what protected the field's analysis against the persistence of the wrong conclusions. Identify what would have been required to reach the INSAG-7 conclusions in 1986.
\end{exercise}

\begin{exercise}
A forensic software-engineering investigation reaches the conclusion "the root cause was operator error: the operator deployed the wrong version." Apply the new-view diagnostic moves from Chapter 9 to the conclusion. Identify what the investigation should examine next.
\end{exercise}

\begin{exercise}
Discuss the role of public attention on a major-accident investigation. Argue both sides: "public attention is a protection against capture and inadequate investigation" versus "public attention pressures investigations to issue premature conclusions and degrades the analysis." Identify the structural conditions that determine which side dominates in a given case.
\end{exercise}

\begin{exercise}
A forensic engineer's report concludes "the probable cause was X; contributing factors were Y and Z" with no explicit identification of the parties responsible for X, Y, or Z. Discuss whether the omission is appropriate. Identify the circumstances under which the report should and should not explicitly attribute responsibility.
\end{exercise}

\begin{exercise}
Compare the NTSB report on a major aviation accident (Eastern 401 \cite{acc:ntsb-aar73-14}, Aloha 243 \cite{acc:ntsb-aar94-04}, Air France 447 \cite{acc:bea-af447}, or analogous) with a peer-reviewed academic case study of the same accident. Identify the analytical moves the academic study performs that the official report does not, and vice versa. Discuss what the comparison implies about the complementary roles of official investigation and academic safety engineering.
\end{exercise}
