\chapter{Combustion, explosives, energetic materials}
\label{vol05:ch12}

\epigraph{Combustion is a rapid exothermic chemical reaction between
a fuel and an oxidiser, sustained by the heat it releases. The
distinction between a deflagration and a detonation is whether the
reaction zone moves at subsonic or supersonic speed with respect to
the unreacted material ahead of it. The engineering use of energetic
materials is the engineering of which side of that boundary the
reaction stays on.}{Adapted from S.\ R. Turns, working scope of
combustion in engineering \cite[ch.~1, pp.~1--8]{text:turns-combustion}.}

\chapmeta{Half-life: durable. Archetypes: stability, failure. Exercise target: 23. Project track: physical measurement (laminar flame speed of a candle). Hazard class: medium (open flame, controlled fuel, ventilated room; fire extinguisher at hand). Reader path default: standard, with sections explicitly tagged. Named cases: Bhopal disaster (1984) referenced for runaway reaction context; Imperial Sugar dust explosion (2008) and BP Texas City refinery explosion (2005) for energetic-material-system failures; energetic-material applications described by mechanism only for in-service systems.}

\noindent
A working engineer treats fuel as a stored chemical potential and
combustion as the controlled release of that potential at a rate
matched to the application: slow and steady in a residential furnace,
fast in an automobile cylinder, microsecond-scale in a detonator, sub-
microsecond in a high-explosive. The chapter develops the thermochemistry
that fixes the heat release, the laminar and turbulent flame structures
that propagate it, the Chapman-Jouguet condition that bounds detonation
speeds, the engineering classes of explosives and propellants, the
pyrotechnic compositions that produce light and sound and time-delays,
and the three failure modes (dust explosions, BLEVE, chain reactions)
that recur in process-industry accidents. The chapter closes the
volume: matter chosen, processed, and burned.

\section{Combustion thermochemistry}\pathtag{core}

A combustion reaction is the rapid oxidation of a fuel:
\[
\text{fuel} + \nu_{\text{ox}} \text{O}_2 \to
\nu_{\text{CO}_2} \text{CO}_2 + \nu_{\text{H}_2\text{O}} \text{H}_2\text{O}
+ \cdots,
\]
with stoichiometric coefficients fixed by element balances. For methane
($\text{CH}_4$):
\[
\text{CH}_4 + 2\,\text{O}_2 \to \text{CO}_2 + 2\,\text{H}_2\text{O}.
\]
Each mole of methane combines with two moles of oxygen; the products
are one mole of carbon dioxide and two moles of water. The reaction is
exothermic; the heat released per mole of fuel is the
\engterm{enthalpy of reaction} $\Delta H_{\text{rxn}}$, computed from
the heats of formation of products and reactants:
\[
\Delta H_{\text{rxn}} = \sum \nu_p \Delta H_f^\circ(\text{products}) -
\sum \nu_r \Delta H_f^\circ(\text{reactants}).
\]
For methane combustion to gaseous products (the lower heating value,
LHV), $\Delta H_{\text{rxn}} \approx -802\,\text{kJ}/\text{mol}$; to
liquid water (the higher heating value, HHV), $-890\,\text{kJ}/\text{mol}$.
The difference is the latent heat of condensation of the product
water; engineering combustion calculations use the LHV when the
product water leaves as vapour (almost always) and the HHV only for
condensing-mode appliances or for fuel-cell accounting
\cite[ch.~2]{text:turns-combustion}.

\engterm{Air-fuel ratio and stoichiometry}. Practical combustion uses
air rather than pure oxygen; air is approximately $21\,\%$ O$_2$ and
$78\,\%$ N$_2$ by volume, with the balance argon and trace species.
The molar nitrogen-to-oxygen ratio is $\sim 3.76$. The stoichiometric
air requirement for methane is $2 \times 4.76 = 9.52$ moles of air per
mole of CH$_4$, or, by mass, $17.2\,\text{kg air}/\text{kg fuel}$. The
working dimensionless ratio is the \engterm{equivalence ratio}
$\phi = (F/A)_{\text{actual}} / (F/A)_{\text{stoich}}$: $\phi = 1$ is
stoichiometric, $\phi < 1$ is fuel-lean (excess air), $\phi > 1$ is
fuel-rich (excess fuel). Working ranges: a residential gas burner runs
$\phi \approx 0.9$ -- $0.95$ (slightly lean); a spark-ignition engine
at part load runs $\phi \approx 1.0$ (stoichiometric, for three-way
catalyst operation); a diesel runs $\phi \approx 0.2$ -- $0.5$ (very
lean, overall, with locally rich pockets at the spray cores).

\engterm{Adiabatic flame temperature}. If the heat released by the
reaction is fully retained by the product gases (no heat loss), the
products reach the \engterm{adiabatic flame temperature} $T_{\text{ad}}$.
For stoichiometric methane-air, $T_{\text{ad}} \approx 2226\,\si{\kelvin}$
at standard conditions, computed from $\sum \nu_p c_p \,\mathrm{d}T =
-\Delta H_{\text{rxn}}$ \cite[ch.~2]{text:turns-combustion}. Real
flames lose heat to the surroundings (convection, radiation) and
operate well below $T_{\text{ad}}$; the gap is the working source of
combustion efficiency calculations.

\engterm{Equilibrium products and dissociation}. At $T > 2000\,
\si{\kelvin}$, the products of combustion do not equilibrate at the
nominal end-state composition. CO$_2$ partially dissociates to CO and
O$_2$; H$_2$O partially dissociates to OH and H$_2$; nitrogen
partially oxidises to NO and NO$_2$. The equilibrium product
composition is obtained by minimising the Gibbs free energy of the
gas mixture at the chosen $T$, $p$. Working tools: the NASA CEA code,
the Cantera library, the Chemkin family of solvers. The dissociation
limits the realised flame temperature (the products store some of the
heat in the form of dissociated species' enthalpy) and produces the
working pollutant emissions (CO from incomplete combustion, NO$_x$
from thermal-NO formation through the Zeldovich mechanism).

\engterm{Lower and higher heating values, working numbers}. The
energy density of common engineering fuels (LHV, MJ/kg):

\begin{itemize}[leftmargin=*]
\item Hydrogen H$_2$: $120\,\text{MJ}/\text{kg}$ (highest specific
  energy of any chemical fuel).
\item Methane CH$_4$: $50\,\text{MJ}/\text{kg}$ (natural gas, $\sim
  90\,\%$ methane).
\item Gasoline (representative C$_8$H$_{18}$): $44\,\text{MJ}/\text{kg}$.
\item Diesel fuel: $42\,\text{MJ}/\text{kg}$.
\item Jet A: $43\,\text{MJ}/\text{kg}$.
\item Ethanol C$_2$H$_5$OH: $27\,\text{MJ}/\text{kg}$ (lower; the OH
  is already partially oxidised).
\item Coal (bituminous): $25$ -- $30\,\text{MJ}/\text{kg}$ (variable by
  rank).
\item Wood (oven-dry): $\sim 18\,\text{MJ}/\text{kg}$.
\item Solid rocket propellant (composite, $70\,\%$ AP, $14\,\%$
  Al, balance binder): $9\,\text{MJ}/\text{kg}$ (lower than
  hydrocarbon fuels because the oxidiser is bundled in).
\item TNT: $4.6\,\text{MJ}/\text{kg}$.
\item Lithium-ion battery (chemistry-dependent): $0.5$ -- $0.9\,
  \text{MJ}/\text{kg}$.
\end{itemize}

The ratio of hydrocarbon-fuel energy density to lithium-ion-battery
energy density is $\sim 50$ -- $80$: the working reason aviation has
not converted to electric propulsion at scale as of 2025.

\engterm{Energetic-material energy density}. TNT is a useful working
reference because its energy density ($4.6\,\text{MJ}/\text{kg}$) is
well below that of hydrocarbon fuels. The distinguishing property of
an explosive is not its energy density but its rate of energy release:
TNT releases its $4.6\,\text{MJ}/\text{kg}$ in microseconds, while
gasoline releases its $44\,\text{MJ}/\text{kg}$ in milliseconds (in
an engine cylinder) to seconds (in a fire). The engineering
distinction is rate, not energy.

\begin{principle}[Combustion is the controlled release of stored
chemical potential; the engineering control is on the rate]
A fuel-oxidiser pair has a fixed enthalpy of reaction set by the
chemistry. The engineering use of the pair is the engineering of the
rate at which the enthalpy is released. The same hydrocarbon fuel
burns slowly in a household furnace (kilowatts), quickly in an engine
cylinder (megawatts), and in a fraction of a millisecond in a
detonator (gigawatts). The chapter's working categories (combustion,
deflagration, detonation, explosion, pyrotechnic effect) are
categories of rate, not of energy.
\end{principle}

\section{Flame structure and propagation}\pathtag{core}

A flame is the propagating reaction front of a combustion process.
In its simplest form, a one-dimensional premixed flame, the front
separates unreacted fuel-air mixture ahead from product gases behind;
the reaction zone in between is a thin region (typical thickness
$0.1$ -- $1\,\text{mm}$ at atmospheric conditions) where the chemistry
runs.

\engterm{The laminar premixed flame}. The flame propagates into the
unreacted mixture at the \engterm{laminar flame speed} $S_L$, a
material property of the mixture set by its chemistry, transport, and
temperature. The working values for stoichiometric mixtures with air
at standard conditions: methane $S_L \approx 0.38\,\text{m}/\text{s}$;
gasoline $S_L \approx 0.43\,\text{m}/\text{s}$; hydrogen $S_L \approx
2.5\,\text{m}/\text{s}$ (the highest of any common fuel-air mixture,
which is why hydrogen burns are uniquely fast and risky); acetylene
$S_L \approx 1.5\,\text{m}/\text{s}$.

The flame speed varies with equivalence ratio, with a peak slightly
on the rich side ($\phi \approx 1.05$ -- $1.1$ for hydrocarbons) and
decreasing on both sides. At the flammability limits (lean and rich)
the flame speed approaches zero; the limits set the practical
combustibility range of a fuel-air mixture. Working flammability
limits for methane in air at standard conditions: $5\,\%$ (lean) to
$15\,\%$ (rich) by volume; for hydrogen $4\,\%$ to $75\,\%$ (the
unusually broad rich limit of hydrogen is the engineering hazard).

The scaling relation
\[
S_L \sim \sqrt{\alpha_{\text{th}} \omega},
\]
with $\alpha_{\text{th}}$ the thermal diffusivity of the gas and
$\omega$ the volumetric reaction rate, captures the working
dependence: the flame speed is set by the balance between heat
conduction (which preheats the unreacted gas to ignition temperature)
and chemical reaction rate (which converts the preheated gas to
products). The product $S_L^2 / \alpha_{\text{th}}$ is a working
estimate of the reaction rate.

\engterm{Flame thickness}. The flame's characteristic thickness is
$\delta_L \sim \alpha_{\text{th}}/S_L$. For methane-air at standard
conditions: $\alpha_{\text{th}} \approx 2 \times 10^{-5}\,\text{m}^2/
\text{s}$ and $S_L \approx 0.38\,\text{m}/\text{s}$, so $\delta_L
\approx 50\,\si{\micro\meter}$. The reaction zone is even thinner;
the preheat zone, where the temperature rises but the chemistry has
not yet started, is the bulk of the flame thickness.

\engterm{Diffusion flames}. When fuel and oxidiser are not premixed
but meet at a boundary (a candle, a kitchen-gas burner with the air
shutter closed, a hydrocarbon fire), the flame structure is set by
diffusion of fuel and oxidiser to the reaction sheet. The reaction
sheet is at the stoichiometric mixture fraction; the temperature peaks
near the sheet. Diffusion flames are slower than premixed flames at
the same fuel; the flame speed is not a material property of the fuel
alone but depends on the geometry of the fuel-oxidiser supply.
Diffusion flames are the working flames of pool fires, industrial
furnaces, jet engines, diesel engines (where the fuel spray injects
into compressed air and burns at the spray surfaces), and most
practical combustion devices.

\engterm{Turbulent flames}. At Reynolds numbers above $\sim 2000$ the
flame is turbulent; the turbulent eddies wrinkle the reaction sheet,
increase its surface area, and raise the effective flame speed by
factors of $5$ -- $50$ over the laminar value. The
\engterm{turbulent flame speed} $S_T$ scales approximately as $S_L
(1 + u'/S_L)$, with $u'$ the turbulent velocity fluctuation. Turbulent
combustion is the working regime for engineering applications:
internal-combustion engines, gas turbines, industrial furnaces.

\engterm{Ignition and quenching}. A flame is initiated when a heat
source raises the local fuel-air mixture above its ignition
temperature; the working ignition energy depends on the mixture
composition, the local temperature, and the size of the kernel. The
\engterm{minimum ignition energy} (MIE) for methane-air at $\phi = 1$
is $\sim 0.3\,\text{mJ}$; for hydrogen-air at $\phi = 1$, $\sim 0.02\,
\text{mJ}$ (an order of magnitude lower, the engineering reason
hydrogen ignition by static electricity is a working hazard). The
\engterm{quenching distance} is the smallest gap through which a
flame can propagate without losing too much heat to the walls; for
methane-air, the quenching distance is $\sim 2\,\text{mm}$. Flame
arresters (engineered devices that allow gas flow but quench any
flame propagating against the flow) work by reducing the effective
gap below the quenching distance.

\section{Detonation vs deflagration}\pathtag{core}

A reaction front that propagates at subsonic speed (laminar or
turbulent flames) is a \engterm{deflagration}. A reaction front that
propagates at supersonic speed is a \engterm{detonation}. The
distinction is mechanical, not just chemical: a deflagration is
pressure-equilibrating (the reaction zone runs into nearly atmospheric
pressure unreacted mixture); a detonation is shock-driven (the
reaction zone is preceded by a shock wave that compresses and heats
the unreacted mixture above its ignition temperature, and the
exothermic chemistry sustains the shock).

\engterm{Chapman-Jouguet detonation}. In the idealised one-dimensional
analysis of D.\ L.\ Chapman (1899) and E.\ Jouguet (1905) summarised
in the working combustion-theory texts \cite[ch.~6]{text:williams-combustion},
the detonation propagates at the speed at which the reactants are
shock-compressed, react to completion, and the product gas is moving
at the local speed of sound relative to the wave (the Chapman-Jouguet
condition). The detonation speed for stoichiometric methane-air at
standard conditions is $\sim 1800\,\text{m}/\text{s}$ (Mach $\sim
5.5$ relative to ambient); for stoichiometric hydrogen-oxygen $\sim
2820\,\text{m}/\text{s}$; for TNT in solid form $\sim 6900\,\text{m}/
\text{s}$; for the working high explosive C-4 $\sim 8100\,\text{m}/
\text{s}$; for HMX $\sim 9100\,\text{m}/\text{s}$.

\engterm{The Zeldovich-von Neumann-Doering (ZND) structure}. The
detonation front, on closer examination, is a shock wave followed by
a chemical induction zone (where the shock-compressed reactants are
above the ignition temperature but have not yet reacted) and then a
reaction zone (where the chemistry runs and heat is released). The
ZND model predicts the working detonation speed, the pressure profile
(the von Neumann spike at the shock front, with peak pressures
$\sim 2 \times$ the CJ pressure), and the thickness of the reaction
zone. Working CJ pressures: TNT $\sim 21\,\text{GPa}$, C-4 $\sim
26\,\text{GPa}$, HMX $\sim 40\,\text{GPa}$.

\engterm{Deflagration-to-detonation transition (DDT)}. A confined
deflagration can accelerate by turbulent feedback (the heated and
expanding products push the unreacted gas ahead, the resulting flow
becomes turbulent, the turbulent flame is faster, the heating is
greater, the acceleration continues) and transition to a detonation
under the right geometric and chemical conditions. The DDT distance
depends on the fuel, the confining geometry, and the obstacle pattern
in the unreacted gas. The DDT is the working hazard in long enclosed
fuel-air mixtures: a leaking natural-gas pipeline in a long covered
trench, a hydrogen leak in a long industrial gallery, a propane leak
in a long warehouse. A deflagration that DDTs produces a pressure
load $5$ -- $10\times$ higher than the bounded deflagration alone.

\engterm{Engineering control of detonation versus deflagration}. The
working engineering controls:

\begin{itemize}[leftmargin=*]
\item Ventilation: keep the unreacted fuel-air mixture below the
  lower flammability limit. Gas detectors and ventilation interlocks
  do this in industrial process plants.
\item Confinement control: avoid long enclosed volumes filled with
  flammable mixtures. The petroleum-refining industry's "blowdown"
  practice (a flare stack to which over-pressure gases are routed)
  is the canonical engineering response; the 2005 BP Texas City
  refinery explosion was a failure of this control
  \cite{acc:csb-bp-texas-city-2007}.
\item Inerting: replace the air in a process vessel with nitrogen or
  carbon dioxide so that the fuel cannot reach a flammable mixture
  with oxygen.
\item Flame arresters and detonation arresters: engineered devices
  that quench a propagating flame or detonation by heat loss to the
  arrester walls.
\item Suppression and explosion vents: engineered weak panels that
  release the deflagration pressure before it reaches a destructive
  level, and chemical-suppression systems that inject a flame-
  inhibiting agent into the developing deflagration.
\end{itemize}

\section{Engineering explosives and propellants overview}\pathtag{standard}

\engterm{Explosives}. The working engineering classes:

\begin{itemize}[leftmargin=*]
\item \engterm{Primary explosives}: highly sensitive (to friction,
  impact, heat, spark), used in detonators to initiate secondary
  explosives. Mercury fulminate, lead azide, lead styphnate,
  diazodinitrophenol. Used in microgram-to-milligram quantities;
  almost never in bulk because of sensitivity.
\item \engterm{Secondary explosives}: less sensitive, capable of
  detonation but only when shock-initiated by a primary explosive.
  The bulk explosives of commerce and warfare: TNT (trinitrotoluene),
  HMX, RDX, PETN, tetryl, ammonium nitrate (in suitable form),
  emulsion explosives (water-in-oil emulsions with ammonium nitrate
  as oxidiser).
\item \engterm{Tertiary explosives}: very low sensitivity, requiring
  large shock to detonate. ANFO (ammonium nitrate / fuel oil), the
  working blasting agent of mining and quarrying; cast TNT-based
  compositions; insensitive munitions formulations (IM, with HMX or
  RDX replaced by less sensitive materials such as IMX-101 in modern
  artillery rounds).
\end{itemize}

\engterm{Brisance and the working measures of explosive performance}.
Brisance is the shattering power of an explosive, related to the
pressure of the detonation front and to the rate of energy release.
Working measures:

\begin{itemize}[leftmargin=*]
\item Detonation velocity ($\text{m}/\text{s}$): TNT $\sim 6900$,
  RDX $\sim 8750$, HMX $\sim 9100$, ANFO $\sim 3200$.
\item Detonation pressure (GPa): TNT $\sim 21$, RDX $\sim 34$, HMX
  $\sim 40$, ANFO $\sim 5$.
\item Specific energy (kJ/kg): TNT $\sim 4600$, RDX $\sim 5400$, HMX
  $\sim 5700$, ANFO $\sim 3900$.
\item Sensitivity (impact, friction, electrostatic, shock): measured
  by standard tests (drop-weight impact, friction sensitivity, BAM,
  card-gap shock).
\end{itemize}

The working selection of an explosive is by the four measures: how
fast (velocity), how hard (pressure), how much (energy), how
safe to handle (sensitivity).

\engterm{Propellants}. A propellant is an energetic material designed
to combust in a controlled deflagration mode, not detonation. Two
working classes:

\begin{itemize}[leftmargin=*]
\item \engterm{Gun propellants} (single-base, double-base,
  triple-base): nitrocellulose-based powders that burn in the gun
  chamber to drive a projectile. Working burn rates depend on
  pressure and on grain geometry; the typical gun propellant
  deflagrates at $1$ -- $10\,\text{m}/\text{s}$ at peak chamber
  pressures of $200$ -- $400\,\text{MPa}$.
\item \engterm{Rocket propellants}: liquid (LH$_2$/LOX, RP-1/LOX,
  hydrazine/N$_2$O$_4$) or solid (composite: ammonium-perchlorate
  oxidiser, aluminium fuel, polybutadiene-based binder; or
  double-base: nitrocellulose-nitroglycerin). Working specific
  impulses ($I_{sp}$, in seconds): RP-1/LOX $\sim 300$, LH$_2$/LOX
  $\sim 450$, composite solid $\sim 250$ -- $270$, double-base solid
  $\sim 230$.
\end{itemize}

The engineering distinction between an explosive and a propellant is
the application: an explosive is meant to produce a shock and a
shattering action; a propellant is meant to produce gas at controlled
pressure that does mechanical work. The chemistry of the two classes
overlaps (nitrocellulose is used in both); the engineering is in the
formulation (grain shape, burn rate, sensitivity) that determines
whether the deflagration is sustained as a controlled burn or
accelerates to a detonation.

\section{Pyrotechnics and energetic materials in industry}\pathtag{standard}

\engterm{Pyrotechnic compositions}. A pyrotechnic is an energetic
material designed to produce a specific effect (light, heat, sound,
gas, time delay, ignition) rather than mechanical work. Working
classes:

\begin{itemize}[leftmargin=*]
\item \engterm{Illuminating}: magnesium-based flares (military
  illumination, marine distress, fireworks); chlorate-based stars
  (fireworks, with metal salts giving colour).
\item \engterm{Smoke-producing}: hexachloroethane / zinc oxide
  (HC smoke for screening); coloured smokes for signalling.
\item \engterm{Whistles and sound producers}: potassium picrate /
  potassium chlorate, the sound being generated by oscillating
  pressure waves in a confined burn.
\item \engterm{Delay compositions}: pressed black powder, lead
  thiocyanate-based delays, or barium peroxide / chromium compositions
  for fuse-delay or rocket-staging applications.
\item \engterm{Igniters and primers}: composition matched to ignite a
  specific propellant or explosive at controlled energy; the canonical
  vehicle is the centrefire-cartridge primer (lead styphnate /
  barium nitrate / tetrazene) or the rocket igniter cartridge (boron-
  potassium-nitrate, BKNO$_3$).
\end{itemize}

\engterm{Energetic materials in industry}. Beyond defence and
fireworks, the working industrial uses of energetic materials:

\begin{itemize}[leftmargin=*]
\item \engterm{Blasting in mining and construction}: ANFO and
  emulsions, with detonator-and-cord initiation systems; the working
  basis of essentially all open-pit mining and tunnelling.
\item \engterm{Automotive airbag inflators}: solid-propellant gas
  generators using sodium azide (older designs) or guanidine-nitrate
  formulations (modern). The inflator must produce $\sim 70\,\text{L}$
  of gas in $\sim 30\,\text{ms}$ at controlled temperature and
  composition (the working hazard: a defective inflator can rupture
  and project metal fragments).
\item \engterm{Seat-belt pre-tensioners and ejection-seat
  cartridges}: pyrotechnic actuators that produce controlled
  mechanical work in the millisecond timescale.
\item \engterm{Welding and metal cutting}: thermite (aluminium/iron-
  oxide redox) for rail-track welding and for welding underwater;
  oxyacetylene cutting (a chemical, not a pyrotechnic, but the working
  example of controlled combustion at metal-cutting energy density).
\item \engterm{Aerospace separation and staging}: linear-shaped
  charges that cut metal structure cleanly; explosive bolts that
  separate stages on command; pyrotechnic valves in spacecraft.
\end{itemize}

\engterm{Regulatory regime}. Energetic materials in the EU are
regulated under the REACH and CLP regulations; in the US under the
ATF (Bureau of Alcohol, Tobacco, Firearms and Explosives) for
acquisition and storage, OSHA for workplace use, and DOT for
transport. Working safety standards: NFPA 495 (explosives processing),
ASTM E2491 (additive-manufacturing energetic materials), MIL-STD-1316
(fuze design). The working hazard classification system is the UN
hazard-class system (1.1 mass-detonation, 1.2 projection, 1.3 fire and
moderate projection, 1.4 small fire and projection, 1.5 very insensitive,
1.6 extremely insensitive).

\section{Failure: dust explosions, BLEVE, chain reactions}\pathtag{core}

Three failure patterns are documented across process-industry
accidents. The chapter's failure section names the mechanisms and the
canonical cases.

\engterm{Dust explosions}. A combustible dust (sugar, flour, coal,
metal powder, plastic powder, wood dust) suspended in air at a
concentration above the lower flammable limit can ignite from a small
ignition source (a static spark, a hot bearing, a welding spark) and
develop a deflagration that, in confined geometry, transitions to a
destructive explosion. The mechanism is the same as a gas-phase
deflagration but with a particulate fuel; the particle-air mixture
is the canonical example of a heterogeneous combustion system.

The 2008 Imperial Sugar refinery explosion at Port Wentworth, Georgia
\cite{acc:csb-imperial-sugar-2009} killed $14$ workers and injured
$36$ in a sugar-dust explosion. The Chemical Safety Board investigation
identified the immediate cause as a primary explosion in an enclosed
sugar conveyor; the primary explosion stirred up accumulated sugar
dust on overhead structures throughout the plant, and the resulting
secondary explosions destroyed the facility. The canonical lessons:
combustible dust on overhead structures is a hazard equivalent to
flammable gas in a confined volume; engineering housekeeping (regular
removal of dust accumulation) is the primary protection; explosion
venting and isolation valves on enclosed conveyors are the secondary
protections; ignition-source control alone is inadequate because the
primary ignition is unpredictable.

\engterm{The pentagon of dust explosion}. The working precondition
list for a dust explosion: combustible dust, dispersion in air,
oxidant (oxygen), confinement (for amplification to destructive
pressure), ignition source. The remediation removes one or more
preconditions: dust collection and housekeeping (removes the dispersed
dust), inerting (removes the oxidant), explosion vents and weak panels
(remove the confinement), ignition-source control (removes the
ignition).

\engterm{Boiling-Liquid Expanding-Vapour Explosion (BLEVE)}. A
pressurised vessel containing a liquid at a temperature above its
atmospheric-pressure boiling point, suddenly ruptured (typically by
fire-induced weakening of the shell, or by impact), produces an
explosive vaporisation of the liquid as the pressure drops. The
vapourisation absorbs the latent heat from the liquid mass, which is
typically large; the resulting vapour cloud expands rapidly and, if
flammable, ignites in a fireball that radiates intense thermal
energy. Working examples: propane-tank BLEVEs (the LPG industry's
canonical hazard, with destructive thermal radiation to several
hundred metres for a typical $30\,\text{t}$ propane tank); refrigerant-
truck and rail-tanker BLEVEs.

The mechanism: the fire-weakened shell ruptures; the saturated liquid
inside flashes to vapour as the pressure drops from the saturation
pressure to atmospheric; the vapour expansion launches the shell
fragments at hundreds of metres per second; the vapour cloud ignites
on the fire that caused the rupture; the resulting fireball at $\sim
1000\,\degC$ radiates thermal flux at $200$ -- $400\,\text{kW}/\text{m}^2$
to receptors within a few hundred metres. The engineering response is
fire-protection of the vessel shell (a sprinkler system, an insulation
jacket, an automatic deluge), routing to prevent flame impingement,
and standoff distances in plant layout that account for the BLEVE
fireball envelope.

\engterm{Chain reactions and runaway reactions}. A chain reaction is
a reaction in which an intermediate species produced by the reaction
is consumed in a step that produces more of the intermediate. If each
intermediate produces, on average, more than one additional
intermediate, the reaction is auto-catalytic and can accelerate
explosively. The working examples in process industry:

\begin{itemize}[leftmargin=*]
\item \engterm{Auto-oxidation of organic compounds}: peroxide chain
  reactions in stored ethers, in deteriorating polymer stabilisers,
  in lubricating oils at elevated temperature. The working hazard:
  diethyl ether stored in air for years develops explosive peroxide
  concentrations.
\item \engterm{Thermal runaway in exothermic reactions}: the reaction
  produces heat, the heat raises the temperature, the temperature
  accelerates the reaction (Arrhenius), the reaction produces more
  heat. Without adequate cooling, the system runs away to a
  catastrophic state. The 1984 Bhopal disaster (a runaway methyl-
  isocyanate hydrolysis reaction in a storage tank that released
  $\sim 40\,\text{t}$ of toxic gas \cite{acc:icmr-bhopal-2004})
  killed thousands; the runaway was triggered by water ingress to
  the storage tank, and the engineered cooling system was inadequate
  to handle the resulting reaction.
\item \engterm{Nuclear chain reactions}: the canonical chain reaction
  whose engineering control is the basis of every commercial reactor
  in service. Volume IX develops the reactor-engineering side.
\end{itemize}

\engterm{The 2005 BP Texas City refinery explosion}
\cite{acc:csb-bp-texas-city-2007} killed $15$ workers and injured
$180$ at the isomerisation unit of the BP refinery. The Chemical
Safety Board investigation identified the immediate cause as
overflow of the raffinate splitter tower with hydrocarbon discharge
to atmosphere through a blowdown stack, ignition of the resulting
vapour cloud by an idling truck, and a vapour-cloud explosion that
destroyed temporary trailers in the unit. The canonical lessons: a
process upset (a level instrument failure, a startup-procedure
deviation) that releases combustible hydrocarbon to atmosphere is the
recurring failure mode in petroleum refining; modern practice uses
flare stacks rather than open blowdown stacks; occupied buildings are
relocated outside the vapour-cloud-explosion envelope; the cultural
contribution (process-safety culture deficiencies identified by the
CSB) is the working organisational lesson.

\engterm{The common structure of energetic-material failures}.
Across the three patterns, the structure is the same: a combustible
or reactive material, a release mechanism (containment failure, dust
accumulation, runaway reaction), an ignition source, and a confining
or amplifying geometry. The engineering remediation, on each of the
four elements, is the working basis of process-safety engineering.

\begin{principle}[Energetic materials in service operate by an
engineered rate; the failure mode is a rate the engineering did not
account for]
A controlled combustion in a furnace, an engine cylinder, or a rocket
motor produces work at a rate the design accommodates. A failed
combustion (dust explosion, BLEVE, vapour-cloud explosion, runaway
reaction) produces the same energy release at a higher rate, in a
different geometry, against a structure that was not designed for
that load. The engineering of energetic materials is the engineering
of bounds on the realisable rates: keep the mixture below flammability,
keep the volume below confinement, keep the temperature below ignition,
keep the cooling above runaway. Where the bounds fail, the failure is
the failure mode.
\end{principle}

\begin{archetype}[Stability and failure]
A combustion system is a dynamical system whose stability depends on
the balance between heat generation (the reaction rate) and heat loss
(conduction, radiation, mass transfer). A stable flame sits at the
balance point. A runaway combustion is the same system with heat
generation exceeding heat loss; the temperature rises, the reaction
accelerates, the heat generation rises further, the cycle runs to a
destructive outcome. The chapter's working examples (dust explosion,
BLEVE, runaway reaction, deflagration-to-detonation transition) are
the failure modes of the same stability problem. The discipline of
the working engineer is to keep the operating point on the stable
side of the balance and to provide the safety devices (flame
arresters, vents, cooling, inerting) that prevent the system from
crossing into the unstable region.
\end{archetype}

\begin{estimation}
\textbf{A typical paraffin candle has a flame approximately $40\,\text{mm}$
tall and $10\,\text{mm}$ wide at its base. Estimate the laminar flame
speed of the candle's diffusion flame and the chemical power released.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The candle's wick wicks liquid paraffin (a long-chain alkane,
$\sim$~C$_{25}$H$_{52}$, $\Delta H_{\text{c}} \approx 43\,\text{MJ}/
\text{kg}$) into the diffusion-flame envelope. The flame is a
non-premixed diffusion flame: vaporised paraffin diffuses outward,
oxygen diffuses inward, the reaction sheet is at the stoichiometric
contour.

The mass-consumption rate of paraffin is, working from typical
candle data, $\sim 0.1\,\text{g}/\text{min}$, or $\sim 1.7 \times
10^{-6}\,\text{kg}/\text{s}$. The chemical power released is
$\dot{Q} = \dot{m} \Delta H_c \approx 1.7 \times 10^{-6} \times 4.3
\times 10^7 \approx 75\,\text{W}$. This is the working answer to one
part of the question.

The propagation speed of a diffusion flame is not a material property
of the fuel alone; it is set by the geometry of fuel-oxidiser delivery.
The candle's flame is a buoyancy-driven jet flame: the hot products
rise at velocity $u_{\text{buoy}} \approx \sqrt{2 g L (\Delta T /
T_{\text{ambient}})}$, with $L \approx 40\,\text{mm}$ the flame
height and $\Delta T / T_{\text{ambient}} \sim 5$ (the flame
temperature is $\sim 1700\,\si{\kelvin}$, ambient is $\sim 300\,
\si{\kelvin}$): $u_{\text{buoy}} \approx \sqrt{2 \times 9.81 \times
0.04 \times 5} \approx 2\,\text{m}/\text{s}$. The candle's products
rise at $\sim 1$ -- $2\,\text{m}/\text{s}$.

The corresponding hydrocarbon-air premixed flame speed for a
representative alkane is $\sim 0.4\,\text{m}/\text{s}$ at $\phi = 1$
(somewhat lower than methane's flame speed because longer alkanes
have lower flame speeds). This is the working laminar flame speed of
the premixed analogue; the candle's diffusion flame propagates by
diffusion of fuel and oxidiser to the reaction sheet rather than by
propagating into an unreacted mixture, so the flame-speed concept
applies only loosely.

The chapter project develops the working measurement of the
candle's laminar flame speed by image analysis of the flame's
inverted cone (the flame sheet in a stagnation-flow configuration
gives a one-dimensional flame-speed measurement).

The postmortem: the candle is the canonical small-scale combustion
experiment; the $75\,\text{W}$ power output and the $\sim 2\,\text{m}/
\text{s}$ buoyancy-driven flow are the working orders of magnitude.
The diffusion-flame nature of the candle makes the flame-speed
question subtler than for premixed flames; the project develops the
quantitative measurement.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Characterise the laminar flame speed of a candle]
The reader observes and measures the flame of a paraffin candle in
controlled conditions, infers the laminar flame speed from the flame
geometry, and compares with published values for representative
alkane fuels. The deliverable is a fifteen- to twenty-five-page report.

\textbf{Track}. Physical observation and measurement.
\textbf{Hazard class}: medium. Open flame; clear the work area of
flammable materials; have a fire extinguisher within reach.

\textbf{Materials}.

\begin{itemize}[leftmargin=*]
\item A standard paraffin candle of known composition and known wick
  type ($\sim 20\,\text{mm}$ diameter, $\sim 100\,\text{mm}$ length).
\item A digital camera or smartphone capable of recording video at
  $30\,\text{fps}$ or higher.
\item A ruler or scale bar for image calibration.
\item An analytical balance reading to $\pm 0.01\,\text{g}$.
\item Time-measurement tool (smartphone, stopwatch).
\item A draft-free room (close windows, turn off fans).
\end{itemize}

\textbf{Procedure}.

\begin{itemize}[leftmargin=*]
\item Weigh the candle. Light it. Photograph at high resolution
  perpendicular to the flame axis with the ruler in the frame for
  scale.
\item Allow the candle to burn for $10$ -- $20\,\text{min}$ in
  draft-free conditions. Record the burning time precisely. Reweigh
  the candle after extinction. Compute the mass-burning rate
  $\dot{m}$.
\item Compute the chemical power $\dot{Q} = \dot{m} \Delta H_c$
  using $\Delta H_c \approx 43\,\text{MJ}/\text{kg}$ for paraffin.
\item Examine the flame photograph. Identify the bright diffusion-
  flame sheet (the yellow/orange envelope) and the inner blue zone
  (the premixed inner cone, near the wick where vaporised fuel mixes
  with diffused oxygen). Measure the height and width of each region.
\item Estimate the buoyancy-driven flow velocity at the flame top:
  $u_{\text{buoy}} \approx \sqrt{2 g L (\Delta T / T_{\infty})}$.
\item Estimate the laminar flame speed of the premixed inner cone
  from the inner-cone angle $\theta$ and the upward flow velocity:
  $S_L \approx u \sin\theta$.
\end{itemize}

\textbf{Analysis}.

\begin{itemize}[leftmargin=*]
\item Tabulate the measured flame geometry, the mass-burning rate,
  and the inferred power and laminar flame speed.
\item Compare the measured $S_L$ with published laminar flame speeds
  for alkane-air mixtures (Turns, Williams, or other working
  combustion text).
\item Discuss the discrepancies and the limits of the
  inverted-cone-angle method for measuring $S_L$.
\end{itemize}

\textbf{Reflection ($1{,}500$ to $2{,}500$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
\item What is the dominant assumption of the inverted-cone-angle
  method, and how is it violated by the candle's diffusion-flame
  structure?
\item How would the measurement change if the candle were burning in
  a microgravity environment?
\item What is the implied $\phi$ at the inner-cone base where the
  flame speed is measured?
\item One feature of the flame structure that the photograph revealed
  and the reader had not anticipated.
\end{itemize}

\textbf{Deliverable}. The high-resolution photographs of the flame
with scale bar, the mass-burning data, the geometry measurements,
the analysis, and the report.

\textbf{Time}. Two to four hours of observation and measurement,
four to eight hours of image analysis, four to six hours of write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Compute the stoichiometric air-to-fuel mass ratio for the combustion
of propane (C$_3$H$_8$) in air. State the equivalence ratio at which
a propane burner running at $20\,\%$ excess air operates.
\end{exercise}

\begin{exercise}
Compute the adiabatic flame temperature for stoichiometric
methane-air combustion at standard conditions, assuming complete
combustion to CO$_2$ and H$_2$O(g) and ignoring dissociation.
\end{exercise}

\begin{exercise}
A natural-gas-fired furnace has a chemical power input of $50\,
\text{kW}$ and produces flue gases at $250\,\degC$. Compute the
efficiency, given that the LHV of natural gas is $50\,\text{MJ}/
\text{kg}$ and the stoichiometric air-fuel ratio is $17.2$.
\end{exercise}

\begin{exercise}
Compute the laminar flame speed of stoichiometric methane-air at
$1\,\text{atm}$ from the scaling $S_L \sim \sqrt{\alpha_{\text{th}}
\omega}$, given $\alpha_{\text{th}} = 2 \times 10^{-5}\,\text{m}^2/
\text{s}$ and a reaction rate $\omega = 4 \times 10^3\,\text{s}^{-1}$.
\end{exercise}

\begin{exercise}
Compute the detonation pressure of TNT using the working Chapman-
Jouguet relation $P_{\text{CJ}} = \rho_0 D^2 / (\gamma + 1)$, with
$\rho_0 = 1640\,\text{kg}/\text{m}^3$, $D = 6900\,\text{m}/\text{s}$,
and $\gamma = 3$ (typical effective gamma for detonation products).
\end{exercise}

\begin{exercise}
A dust explosion in a sugar refinery has a peak overpressure of
$0.3\,\text{MPa}$ in a $1{,}000\,\text{m}^3$ confined enclosure.
Compute the implied total energy released and compare with the energy
content of $50\,\text{kg}$ of suspended sugar dust.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the equivalence-ratio definition $\phi = (F/A)_{\text{actual}}
/ (F/A)_{\text{stoich}}$ and show that for a given fuel, the flame
speed peaks at $\phi \approx 1.05$ -- $1.1$ rather than at exactly
$\phi = 1$.
\end{exercise}

\begin{exercise}
Derive the adiabatic-flame-temperature equation $\sum \nu_p c_p
\,dT = -\Delta H_{\text{rxn}}$ from the first law applied to a
constant-pressure reacting system.
\end{exercise}

\begin{exercise}
Derive the Chapman-Jouguet condition for a one-dimensional
detonation wave, showing that the product gas velocity (in the
wave frame) equals the local sound speed.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the chemical power output of a residential natural-gas
furnace heating a $200\,\text{m}^2$ house in a temperate climate.
\end{exercise}

\begin{exercise}
Estimate the deflagration overpressure in a $50\,\text{m}^3$ confined
room filled with stoichiometric propane-air, ignited centrally.
\end{exercise}

\begin{exercise}
Estimate the BLEVE fireball diameter and radiated thermal flux at
$500\,\text{m}$ from a fully-flashed $30\,\text{t}$ propane storage
tank.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a Cantera or equivalent equilibrium solver to compute the
adiabatic flame temperature and the equilibrium product composition
for stoichiometric methane-air, hydrogen-air, and propane-air at
$1\,\text{atm}$. Compare the dissociation losses for the three
fuels.
\end{exercise}

\begin{exercise}
Implement a one-dimensional laminar premixed flame solver and
compute the laminar flame speed of methane-air at $\phi = 1.0$,
$1.05$, $1.10$, and $1.20$. Identify the peak flame speed and
compare with published values.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify the ventilation, inerting, and explosion-vent design for a
$200\,\text{m}^3$ sugar-processing room handling $1\,\text{t}/\text{h}$
of granulated sugar. State the dust-control measures, the
explosion-vent area, and the inspection regime.
\end{exercise}

\begin{exercise}
Specify the fire-protection design for a $30\,\text{t}$ liquefied-
petroleum-gas storage tank in a residential-adjacent industrial site.
State the deluge system, the standoff distances, and the bunding.
\end{exercise}

\begin{exercise}
Specify the safety system for an automotive airbag inflator: state the
gas-generator chemistry, the burst-disc rating, the bag-deployment
volume, and the timing.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A natural-gas-fired industrial furnace produces a CO emission of
$500\,\text{ppm}$ in the flue gas. Diagnose the operating condition
(equivalence ratio, mixing, flame stability) that produces the CO
level, and propose a single change to the burner design or the
operating point that would halve the CO.
\end{exercise}

\begin{exercise}
A gasoline-engine emissions test shows higher NO$_x$ at full load
than at part load. Diagnose the combustion-temperature-driven cause
and propose two engineering interventions (EGR, lean-burn,
post-treatment) that address the issue at different points in the
combustion cycle.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A pneumatic-conveyor dust explosion in a wood-pellet plant injures
three workers. The explosion path traces from the dust collector
through the conveyor to the storage silo. Identify the working
preconditions of the pentagon of dust explosion and recommend two
engineering interventions that would have broken the propagation
chain.
\end{exercise}

\begin{exercise}
A storage tank of an industrial solvent (di-isopropyl ether) explodes
during a routine sampling operation. The plant has no record of recent
peroxide testing. Identify the likely failure mode and recommend the
inspection-and-monitoring regime that would prevent recurrence.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A start-up's pitch claims a hydrogen-fuelled passenger-car internal-
combustion engine with comparable safety to gasoline, on the argument
that hydrogen disperses quickly upon leakage. Write a one-paragraph
reply identifying the engineering considerations (broad flammability
limits, low minimum ignition energy, hydrogen embrittlement of
storage materials) that the safety analysis must address.
\end{exercise}

\begin{exercise}
A facility manager proposes to relocate a control room into the
explosion envelope of a hydrocarbon-processing unit on cost grounds,
arguing that the building's reinforced construction provides adequate
protection. Write a one-paragraph evaluation, identifying the
working hazard analyses (vapour-cloud-explosion, BLEVE thermal
radiation, fragment projection) that the proposal must address.
\end{exercise}
