\chapter{Ceramics}
\label{vol05:ch07}

\epigraph{Ceramics are inorganic, non-metallic materials produced by
high-temperature processing. They include the oxides, carbides,
nitrides, borides, sulphides, silicates, glasses, and refractories; they
encompass single crystals and polycrystals, dense bodies and porous
ones, and span hardness from talc to diamond. Their unifying
characteristic is that they fail brittle.}{W.~D. Kingery, H.~K. Bowen,
and D.~R. Uhlmann, paraphrasing the working scope of the field
\cite[ch.~1, pp.~3--5]{text:kingery-ceramics}.}

\chapmeta{Half-life: foundational. Archetypes: failure (brittleness), scaling. Exercise target: 26. Project track: physical measurement. Hazard class: low (chalk and mortar samples; sample preparation involves small hand tools and safety glasses). Reader path default: standard, with sections explicitly tagged. Named cases: Hubble primary mirror (1990) referenced for brittleness in ground-glass optics; concrete spalling cases described by mechanism only.}

\noindent
Ceramics are the working material class whose properties run opposite
to the metals chapter. They are hard, stiff, refractory, and chemically
inert; they fail without warning by brittle fracture. The chapter
develops the microstructural origin of those properties, the Weibull
statistics required to design with a material that has no plastic
reserve, the three engineering ceramic families (glasses, refractories,
concrete), the electronic ceramics that underwrite half the sensor and
actuator inventory in Volume IX, and the three brittle failure modes
the engineer must name before specifying a ceramic part. Volume III
gave the loads; Volume V Chapter 5 gave the metals; this chapter gives
the materials that take the loads metals cannot and refuse to bend.

\section{Ceramic structure and bonding}\pathtag{core}

A \engterm{ceramic} is an inorganic, non-metallic solid in which the
bonding is predominantly ionic, covalent, or a mixture of the two. The
working examples are oxides ($\mathrm{Al_2O_3}$, $\mathrm{SiO_2}$,
$\mathrm{MgO}$, $\mathrm{ZrO_2}$), carbides ($\mathrm{SiC}$,
$\mathrm{WC}$, $\mathrm{B_4C}$), nitrides ($\mathrm{Si_3N_4}$,
$\mathrm{BN}$, $\mathrm{AlN}$), borides ($\mathrm{TiB_2}$), silicates
(clays, micas, feldspars, the bulk of the earth's crust), and the
glasses derived from silicate, borosilicate, and aluminosilicate melts.
The four-volume scope of Kingery, Bowen, and Uhlmann remains the
canonical reference for the structural and processing details
\cite[chs.~2--6]{text:kingery-ceramics}.

The bonding character determines the working properties. Ionic bonds
between cation and anion are non-directional but strong: charges of
opposite sign are pulled together by the Coulomb attraction
$U_{\text{Coulomb}} = -k_e z_1 z_2 e^2/r$, balanced at equilibrium by
the short-range repulsion of overlapping electron shells (the Born
repulsion $\propto 1/r^n$ with $n \sim 9$ for many oxides). The
equilibrium spacing is set by the cation and anion radii; the lattice
energy per mole, in the Born-Land\'e form,
\[
U_{\text{lattice}} = -\frac{N_A M k_e z^+ z^- e^2}{r_0}
\left(1 - \frac{1}{n}\right),
\]
runs to hundreds of kilojoules per mole for representative ionic
ceramics ($U_{\text{lattice}} \approx 3795\,\text{kJ/mol}$ for NaCl,
roughly twice that for the divalent MgO, eight times for the trivalent
$\mathrm{Al_2O_3}$). The lattice energy sets the melting point, the
elastic stiffness, the thermal expansion, and the resistance to
chemical attack.

Covalent bonds between like or near-like-electronegativity atoms are
directional and even stronger. Silicon carbide and boron nitride
inherit the diamond-like tetrahedral coordination from their group-IV
parent; the bond angles are fixed; the bond stretching and bending
energies are large. Covalent ceramics are the hardest engineering
materials known: diamond at $\sim 70\,\text{GPa}$ on Knoop hardness,
cubic boron nitride at $\sim 50\,\text{GPa}$, silicon carbide at
$\sim 25\,\text{GPa}$.

Mixed bonding is the working case for most engineering oxides. The
electronegativity difference between cation and anion in Pauling's
scale measures the ionic fraction; for $\mathrm{Al_2O_3}$ the ionic
fraction is roughly $0.63$, with a covalent contribution that
stiffens the lattice further. The chapter's quantitative claims about
moduli, strengths, and melting points all trace to the
ionic-plus-covalent bonding character.

\engterm{Crystal structures}. The packing of ions in a ceramic lattice
follows two rules: cations and anions touch (the radius-ratio
constraint), and the structure is electrically neutral (the charge
balance). The radius-ratio constraint
$r^+/r^- = \rho$ determines the cation's coordination number: for
$\rho < 0.22$ the coordination is two; for $0.22 < \rho < 0.41$ it is
four; for $0.41 < \rho < 0.73$ it is six; for $\rho > 0.73$ it is eight.
The constraint is geometric and follows from packing a hard sphere
of one radius around hard spheres of another \cite[ch.~2]{
text:kingery-ceramics}.

Working examples:
\begin{itemize}[leftmargin=*]
\item Rock-salt structure (NaCl, MgO, CaO, FeO): face-centred-cubic
  anion sublattice with cations in the octahedral interstices.
  Coordination $6:6$.
\item Fluorite structure ($\mathrm{CaF_2}$, $\mathrm{ZrO_2}$ above
  $2370\,\degC$, $\mathrm{UO_2}$): face-centred-cubic cation
  sublattice with anions in all tetrahedral holes. Coordination $8:4$.
\item Corundum structure ($\mathrm{Al_2O_3}$, $\mathrm{Cr_2O_3}$,
  $\mathrm{Fe_2O_3}$): hexagonal-close-packed anion sublattice with
  cations in two-thirds of the octahedral holes.
\item Perovskite structure ($\mathrm{BaTiO_3}$, $\mathrm{SrTiO_3}$,
  $\mathrm{CaTiO_3}$, $\mathrm{PbZrO_3}$, the high-$T_c$ cuprates):
  large A-cation at the cube corners, small B-cation at the cube
  centre, oxygens at the face centres. The structure family
  underwrites the electronic ceramics of section~7.6.
\item Spinel structure ($\mathrm{MgAl_2O_4}$, $\mathrm{Fe_3O_4}$):
  close-packed oxygen sublattice with cations distributed among
  tetrahedral and octahedral sites.
\end{itemize}

The crystal structure is the lever by which the materials engineer
trades one property for another. Doping a perovskite with a different
A-cation shifts its Curie temperature; substituting an aliovalent
cation in a fluorite introduces oxygen vacancies and increases ionic
conductivity (the working basis for the yttria-stabilised zirconia
electrolyte in solid-oxide fuel cells); replacing an iron in spinel
with a heavier cation tunes its magnetic susceptibility.

\engterm{Defect chemistry}. Ceramics admit the same point-defect
families as metals (vacancies, interstitials, substitutional solutes)
but with the additional constraint of electroneutrality. A cation
vacancy in NaCl must be paired with an anion vacancy (a Schottky
defect) or with a cation interstitial (a Frenkel defect) to keep the
crystal electrically neutral. The equilibrium defect concentration
follows the same Boltzmann form as in metals:
$n_v/N = \exp(-E_v / k_B T)$, with $E_v$ the formation energy of the
defect pair. Defect concentrations control ionic conductivity, diffusion
in solid-state reactions, sintering, and creep.

\begin{archetype}[Failure]
A ceramic part fails by the propagation of a crack from a pre-existing
flaw. The crack does not blunt by plastic deformation; the stress at
the crack tip exceeds the cohesive strength of the bond, and the crack
runs. The chapter's quantitative work, from the Griffith criterion to
the Weibull distribution to the brittle-to-ductile transition, all
follows from the absence of dislocation motion. The engineering use of
ceramics is the engineering management of the flaw population:
inspection finds the largest flaws, surface treatment removes them,
prestress closes them, and proof tests certify their absence above a
chosen size. A ceramic part with no flaws and no surface damage would
fail at the cohesive strength of its bonds, several gigapascals; the
working strength of real parts is one to two orders of magnitude lower.
\end{archetype}

\section{Mechanical behaviour: brittleness, Weibull statistics}\pathtag{core}

A ceramic loaded in tension behaves elastically up to fracture. The
stress-strain curve is a straight line from the origin to the
fracture point; the elastic modulus $E$ runs from $70\,\text{GPa}$
(silica glass) to $400\,\text{GPa}$ (alumina) to $700\,\text{GPa}$
(silicon carbide) to over $1000\,\text{GPa}$ (diamond). The strain at
fracture is small: a few tenths of a percent for most engineering
ceramics, a few percent only for the toughest. There is no yield, no
necking, no plastic reserve. The fracture mode is brittle: the crack
runs at the speed of an elastic wave once it starts.

The cohesive strength of an ideal ionic or covalent bond, estimated
from the Frenkel argument, is approximately $E/10$ to $E/15$: tens of
gigapascals. Real ceramics fracture at one to two hundredths of this
value. The discrepancy is the engineering content of the Griffith
analysis.

\engterm{The Griffith criterion}. A surface crack of length $a$ in a
brittle solid under remote tensile stress $\sigma$ reduces the elastic
strain energy by an amount that scales as $\pi a^2 \sigma^2/E$ per
unit thickness and increases the surface energy by $4 a \gamma_s$ per
unit thickness (two new surfaces of length $a$). The crack propagates
when the differential energy with respect to crack length is positive
\cite[ch.~15]{text:callister2018}, giving the critical stress
\[
\sigma_f = \sqrt{\frac{2 E \gamma_s}{\pi a}}.
\]
The fracture stress falls as the square root of the crack length. A
$10\,\si{\micro\meter}$ crack in alumina ($E \approx 380\,\text{GPa}$,
$\gamma_s \approx 1\,\text{J}/\text{m}^2$) gives
$\sigma_f \approx 150\,\text{MPa}$, an order of magnitude below the
cohesive strength and consistent with the measured working strength.
The criterion is the formal statement that the strength of a ceramic
part is the strength of its largest flaw.

The Griffith analysis generalises to non-ideal cracks through the
\engterm{stress-intensity factor} $K_I = Y \sigma \sqrt{\pi a}$, with
$Y$ a geometry factor of order unity. Fracture occurs when $K_I$
reaches the material's \engterm{fracture toughness} $K_{Ic}$. Working
values: $K_{Ic} \approx 0.7\,\text{MPa}\cdot\sqrt{\text{m}}$ for soda-
lime glass, $\sim 3\text{--}4$ for alumina, $\sim 5$ for silicon
nitride, $\sim 8$ for partially-stabilised zirconia (the
transformation-toughened champion of structural ceramics),
$\sim 50\text{--}100$ for engineering steels. The ceramic toughness
values are an order of magnitude below those of metals; the working
strength is governed by the relation $\sigma_f = K_{Ic}/(Y\sqrt{\pi
a})$ applied to the flaw size detectable by the available
non-destructive inspection.

\engterm{Compression versus tension}. Ceramics carry compression much
better than tension. A Griffith crack under compression does not
extend along its original plane (it tends to close); cracks propagate
under compression only by branching laterally from the existing flaws
at much higher stresses. Compressive strengths of structural ceramics
are typically $8$--$15$ times the tensile strength. The whole design
discipline of ceramics is to keep the working stress compressive: arches
in masonry, prestress in concrete, residual compressive surfaces in
toughened glass.

\engterm{Weibull statistics}. The flaw population in a ceramic is a
random variable, so the strength is a random variable. The
\engterm{Weibull distribution}, introduced by Waloddi Weibull in 1939
\cite{paper:weibull1939}, gives the probability that a ceramic
specimen of volume $V$ survives a tensile stress $\sigma$:
\[
P_s(\sigma) = \exp\!\left[ - \frac{V}{V_0}
\left(\frac{\sigma - \sigma_u}{\sigma_0}\right)^m \right],
\]
where $\sigma_u$ is a threshold stress (often taken zero for
engineering ceramics), $\sigma_0$ is a scale parameter (the stress at
which the probability of survival of unit volume equals $1/e$), and
$m$ is the \engterm{Weibull modulus}. The Weibull modulus measures
the scatter in strength: $m \approx 5$ for chalk and porous bricks,
$m \approx 10$--$15$ for engineering oxides and nitrides,
$m \approx 25$--$40$ for proof-tested silicon carbide turbine parts,
$m \approx 100$ for metals (which fail by plastic mechanisms with
much narrower distributions). A higher $m$ is the engineering goal:
narrower scatter, more predictable failure.

The Weibull distribution gives a working law for the scaling of
strength with volume. The probability that a part of volume $V$ has
no flaw of critical size is the probability that no element of volume
$V_0$ has such a flaw, raised to the power $V/V_0$. Two specimens of
the same material with volumes $V_1$ and $V_2$, tested at the same
probability of failure, have strengths in the ratio
\[
\frac{\sigma_1}{\sigma_2} = \left(\frac{V_2}{V_1}\right)^{1/m}.
\]
A ceramic structural part scaled up by a factor of $10$ in volume
loses approximately $10^{1/m}$ of its working strength: at $m = 10$,
a $20\,\%$ loss; at $m = 5$, a $58\,\%$ loss. The scaling is the
single most consequential consequence of brittle behaviour for
structural ceramics, and it is the reason a ceramic part that passed
a small-coupon test routinely fails in a full-scale assembly.

\begin{principle}[The strength of a ceramic is the strength of its
largest flaw]
A brittle solid fractures at the stress at which an existing crack of
the critical size becomes unstable under the Griffith criterion. The
working strength is set by the flaw population, not by the cohesive
strength of the bond. Halving the largest detectable flaw raises the
working strength by $\sqrt{2}$; doubling the inspected volume reduces
the working strength by $2^{1/m}$. Every engineering decision about a
ceramic part, from material choice to surface preparation to
non-destructive inspection, is a decision about the flaw population.
\end{principle}

\engterm{Fitting a Weibull distribution to measured data}. Strength
data $\{\sigma_i\}$ from $N$ nominally identical specimens are sorted
into ascending order. The empirical cumulative probability of failure
is estimated as $P_{f,i} = (i - 0.5)/N$ or $P_{f,i} = i/(N+1)$
(several plotting-position estimators are in use). The Weibull form
linearises as
\[
\ln \ln \frac{1}{1 - P_{f,i}} = m \ln \sigma_i - m \ln \sigma_0,
\]
so a plot of $\ln\ln[1/(1-P_f)]$ versus $\ln \sigma$ produces a
straight line of slope $m$ and intercept $-m \ln \sigma_0$. The chapter
project takes a stack of chalk sticks (a brittle ceramic with
deliberately wide flaw distribution) and applies the procedure end
to end.

\begin{estimation}
\textbf{A ceramic disc-brake rotor is to be operated in a passenger
car at a maximum brake-line pressure that produces a peak tensile
stress of $200\,\text{MPa}$ at the rotor's friction surface during
heavy braking. The rotor is silicon-carbide-matrix reinforced
($\sigma_0 = 600\,\text{MPa}$ at the laboratory-coupon scale of
$V_0 = 1\,\text{cm}^3$, Weibull modulus $m = 12$) and the working
rotor has a volume of $300\,\text{cm}^3$. What is the probability of
in-service fracture per brake application?}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The scaling factor for the rotor is $V/V_0 = 300$. The working
strength at the same probability of failure as the coupon scales as
$\sigma = \sigma_0 (V_0/V)^{1/m} = 600 \times (1/300)^{1/12}
= 600 \times \exp(-\ln 300 / 12) = 600 \times 0.62 = 370\,\text{MPa}$.

The working stress of $200\,\text{MPa}$ is a fraction
$\sigma/\sigma_0 = 200/600 = 0.333$ of the coupon-scale parameter.
With the volume factor $V/V_0 = 300$ included, the survival probability
per loading event is
\[
P_s = \exp\!\left[-300 \times (200/600)^{12}\right]
= \exp(-300 \times 5.4 \times 10^{-6}) = \exp(-0.0016) \approx 0.998.
\]
The probability of fracture per brake application is approximately
$2 \times 10^{-3}$.

For a passenger car that performs $\sim 10^6$ brake applications over
its design life, this probability is catastrophic: it predicts roughly
$2000$ rotor fractures over the life of the car. Either the working
stress must come down, the Weibull modulus must come up (by proof
testing and screening out the lower tail of the distribution), or the
material chemistry must change. Carbon-ceramic disc-brake rotors in
production, on which our numbers are loosely modelled, run at lower
working stresses and rely on proof tests that censor the lowest tail
of the distribution; the field failure rate is several orders of
magnitude lower than our raw estimate, and the rotors are reserved for
high-performance applications where the cost-and-mass benefit
justifies the testing burden.

The postmortem: the estimate is honest about the volume scaling and
about the Weibull tail, and is therefore brutal about what a structural
ceramic asks of the design discipline. The arithmetic is not pessimism;
it is the price of working with a material that has no plastic reserve.
\end{estimation}

\section{Glasses}\pathtag{standard}

A \engterm{glass} is a non-crystalline solid produced by cooling a
melt fast enough to prevent crystallisation. The atomic arrangement is
amorphous: short-range order is preserved (every silicon in a silicate
glass is bonded to four oxygens), but long-range order is absent. The
working glass families are silicate (soda-lime, the bulk of flat-glass
production), borosilicate (Pyrex, laboratory glassware, low expansion),
aluminosilicate (chemically-strengthened cover glass for displays),
fused silica (laboratory and optical), lead silicate (radiation
shielding, formerly crystal), and the chalcogenide glasses (sulphide,
selenide, telluride: infrared optics).

\engterm{The glass transition}. On cooling, a glass-forming melt
becomes increasingly viscous. At the \engterm{glass transition
temperature} $T_g$, the viscosity reaches $\eta \approx 10^{12}\,
\text{Pa}\cdot\text{s}$; on the timescale of cooling, the structure can
no longer rearrange, and the supercooled liquid freezes into a glass.
The transition is not a phase transition in the thermodynamic sense:
$T_g$ depends on the cooling rate (lower cooling rates give lower
$T_g$). The viscosity follows a Vogel-Fulcher-Tammann form:
\[
\log_{10}\eta = A + \frac{B}{T - T_0},
\]
which is consistent with the thermodynamic Adam-Gibbs picture of
fragility \cite[ch.~3]{text:kingery-ceramics}.

Soda-lime glass has $T_g \approx 550\,\degC$ and a working point
(viscosity $\sim 10^3\,\text{Pa}\cdot\text{s}$, suitable for pressing
or blowing) of about $1000\,\degC$. Fused silica has $T_g \approx
1100\,\degC$ and a working point above $1700\,\degC$; the high working
temperature is why fused silica is processed by chemical vapour
deposition (for optical-fibre preforms) rather than by direct melt
work.

\engterm{The strength of glass and surface flaws}. The cohesive
strength of silica is approximately $E/10 \approx 7\,\text{GPa}$. The
working strength of as-drawn glass fibres is $3$--$5\,\text{GPa}$; the
working strength of plate glass is $30$--$100\,\text{MPa}$, two orders
of magnitude below the cohesive value. The difference is the flaw
population introduced by handling: every contact with another solid
introduces surface scratches; every scratch is a Griffith crack with
length proportional to the scratch depth. The pristine surface of a
freshly drawn fibre survives until the first touch.

The engineering management of glass strength is the management of
surfaces. Three strategies:

\begin{itemize}[leftmargin=*]
\item \engterm{Annealing}: cool slowly through the glass transition to
  relieve residual tensile stresses. Annealed glass has the strength
  of its surface flaw distribution and breaks into long, sharp shards.
\item \engterm{Tempering}: cool the surface fast (air jets, oil
  quench) while the interior is still hot. The contracting interior
  pulls the now-rigid surface into residual compression
  ($-100$ to $-150\,\text{MPa}$ in tempered float glass). Surface
  flaws can grow only when the applied tensile stress exceeds the
  residual compression. Tempered glass has working strengths four to
  five times those of annealed glass and shatters into small,
  cube-edge fragments rather than long shards. Automotive side
  windows, oven doors, and architectural safety glass all rely on
  tempering.
\item \engterm{Chemical strengthening (ion exchange)}: replace small
  sodium ions in the glass surface with larger potassium ions through
  contact with a molten KNO$_3$ bath. The larger ions wedge the
  surface into compression ($-500$ to $-800\,\text{MPa}$ at the
  surface, with a depth of compression of $10$--$60\,\si{\micro\meter}$).
  Chemically strengthened aluminosilicate cover glass on consumer
  electronics relies on this mechanism; the trade-off is high
  surface strength against modest depth, so a deep gouge that
  penetrates the compression layer drops the strength to the
  annealed value or below.
\end{itemize}

The three strategies are listed in order of increasing residual
compression and increasing surface-strength-per-unit-cost. The choice
is set by the application's demand and by the available processing
infrastructure.

\engterm{Optical glasses}. The refractive index $n$ and Abbe number
$V_d$ (a measure of dispersion) span a two-dimensional design space
populated by silicate, phosphate, lanthanum, and fluoride glass
families. Schott's optical-glass catalogue lists hundreds of working
compositions; the optical designer chooses two or three glasses with
chromatic-aberration-correcting combinations of $n$ and $V_d$. The
correlation between density and refractive index (heavier glasses
generally have higher $n$) and between $n$ and $V_d$ (high-index glasses
generally have low $V_d$) sets the working trade-off. Volume IV
Chapter 11 develops the optical side; this chapter notes only that the
glass-selection problem is solved by the same Ashby-style charts that
section~10 of this volume applies to structural materials.

\section{Refractories and high-temperature applications}\pathtag{standard}

A \engterm{refractory} is a non-metallic material that retains useful
mechanical properties at temperatures above approximately $1500\,
\si{\kelvin}$. The working refractories cover the linings of steel
ladles, glass tanks, cement kilns, blast furnaces, and incinerators;
the hot-section components of gas turbines (in single-crystal
nickel-superalloy form, but coated with thermal-barrier ceramics);
the heat shields of reentry vehicles (carbon-carbon, carbon-silicon
carbide); the susceptors and heaters of semiconductor processing
equipment. The temperature regime is where ceramics are mandatory:
metals lose strength above $\sim 0.6 T_m$, and even nickel
superalloys are limited to about $1100\,\degC$ in the gas turbine
working environment.

\engterm{Refractory selection by chemistry}. The basic axis is the
acid-base character of the refractory and of the slag or melt it
contacts. \engterm{Acid refractories} are silica-rich
($\mathrm{SiO_2}$ bricks, fireclay, silica glass) and resist acidic
slags but corrode in basic slags. \engterm{Basic refractories} are
magnesia-rich (magnesite $\mathrm{MgO}$, dolomite, chrome-magnesia)
and resist basic slags but spall in acidic environments.
\engterm{Neutral refractories} (alumina, chromite, mullite, silicon
carbide) bridge the two. The basic-oxygen steelmaking furnace uses a
basic magnesia lining because the slag is lime-rich; the silica glass
tank uses a fused-cast alumina-zirconia-silica lining (AZS) because
the molten glass is acidic.

\engterm{Thermal shock resistance}. A refractory subjected to a sudden
temperature change develops thermal stresses
\[
\sigma_{\text{thermal}} = \frac{E \alpha \Delta T}{1 - \nu},
\]
where $\alpha$ is the linear thermal expansion coefficient. For
alumina ($E \approx 380\,\text{GPa}$, $\alpha \approx 8 \times
10^{-6}\,\si{\per\kelvin}$, $\nu \approx 0.23$), a temperature shock
of $\Delta T = 200\,\si{\kelvin}$ produces $\sigma_{\text{thermal}}
\approx 790\,\text{MPa}$, well above the working tensile strength of
$\sim 300\,\text{MPa}$. The \engterm{thermal shock parameter}
$R = \sigma_f (1 - \nu)/(E \alpha)$ is a working figure of merit:
$R$ for alumina is approximately $75\,\si{\kelvin}$, $R$ for fused
silica (because of its anomalously low expansion coefficient
$\alpha \approx 0.5 \times 10^{-6}\,\si{\per\kelvin}$) is over
$1000\,\si{\kelvin}$.

The high thermal-shock resistance of fused silica is why it is used
for high-temperature laboratory crucibles, certain investment-casting
moulds, and the cores of optical-fibre preforms; the low resistance of
alumina is why alumina kiln furniture must be preheated and cooled on
a controlled ramp; the spalling of refractory linings on rapid cooling
is the engineering failure mode named in section~7.

\engterm{Cordierite, lithium aluminosilicates, and other near-zero-
expansion ceramics}. Cordierite ($\mathrm{Mg_2Al_4Si_5O_{18}}$) has
$\alpha \approx 1.5 \times 10^{-6}\,\si{\per\kelvin}$ and is the
working substrate for automotive catalytic-converter monoliths and
diesel particulate filters: the substrate must survive thermal
cycling between ambient and $\sim 900\,\degC$ on every cold start, and
no other low-cost ceramic survives the cycling. Lithium-aluminosilicate
glass ceramics (LAS, the basis of Schott Ceran and Corning Visions)
have $\alpha \approx 0$ in the working temperature range and are the
working stovetops and high-end telescope mirrors. The mechanism is
microstructural: the negative-expansion ceramic phase and the
positive-expansion glass phase combine in a composite that has
approximately zero net expansion.

\engterm{Refractory composites}. The hot section of a modern gas
turbine uses ceramic-matrix composites (CMCs) of silicon carbide
fibres in a silicon carbide matrix for combustion-liner and turbine-
shroud parts. The CMC trades the brittleness of monolithic SiC for the
toughness of a fibre-bridged composite (Volume V Chapter 8 develops
the mechanism). Operating temperatures of $1300$--$1400\,\degC$ are
achievable in service, against $\sim 1100\,\degC$ for the best metallic
superalloys; the resulting thermodynamic improvement in cycle
efficiency is several percent. The same mechanism underlies the
carbon-carbon and carbon-SiC composites used as reentry-vehicle heat
shields: the leading edges of the Space Shuttle Orbiter were
reinforced-carbon-carbon, and the breach in the Columbia accident's
left-wing leading edge was a foam-impact failure of an RCC panel
\cite{acc:caib-columbia}.

\section{Concrete: the most-used engineered material}\pathtag{core}

\engterm{Concrete} is the engineered ceramic of the built environment:
global production is approximately $4.2\,\text{Gt}$ per year of
cement, current as of 2024, which yields approximately $30\,\text{Gt}$
of concrete \cite[ch.~1]{text:neville-concrete}. By volume placed, it
is the second most consumed material in the world after water. Every
chapter of this book is, more or less, inside a concrete building.

\engterm{Composition and hydration}. Portland cement is a powder
produced by grinding clinker, which is itself the product of
calcining (heating) limestone ($\mathrm{CaCO_3}$) and clay together at
approximately $1450\,\degC$. The clinker contains four principal
phases: tricalcium silicate ($\mathrm{Ca_3SiO_5}$, abbreviated
$\mathrm{C_3S}$, the strength-developing phase); dicalcium silicate
($\mathrm{Ca_2SiO_4}$, $\mathrm{C_2S}$, slow-strength-developing);
tricalcium aluminate ($\mathrm{C_3A}$, sets fastest, controlled by
gypsum addition); and tetracalcium aluminoferrite
($\mathrm{C_4AF}$). On mixing with water, the clinker phases hydrate
to form calcium-silicate-hydrate gel (C-S-H, the strength-bearing
phase) and calcium hydroxide. The hydration reactions are exothermic
and continue for years; the canonical strength-gain curve climbs to
$\sim 70\,\%$ of the 28-day strength in 7 days, $100\,\%$ at 28 days,
and continues to $\sim 130\,\%$ after several years \cite[ch.~4]{
text:neville-concrete}.

\engterm{Mix design and Abrams' law}. The compressive strength of
concrete is inversely related to the water-to-cement ratio
$w/c$. Abrams' law, established empirically by Duff Abrams in 1918
and still the working basis of mix design,
\[
f_c = \frac{A}{B^{w/c}},
\]
with $A$ and $B$ constants for a given cement and aggregate, captures
the relation: doubling the $w/c$ from $0.4$ to $0.8$ drops the
28-day compressive strength from $\sim 50\,\text{MPa}$ to
$\sim 15\,\text{MPa}$. The working trade-off is between strength
($w/c$ low) and workability ($w/c$ high enough to place and consolidate
the mix). Modern superplasticisers (sulphonated melamine or
naphthalene condensates, polycarboxylate ethers) maintain workability
at $w/c$ ratios as low as $0.25$; they are the chemistry that
underwrites the high-strength concretes of contemporary tall-building
construction.

\engterm{Working compressive strengths}. Normal-strength structural
concrete: $25$--$40\,\text{MPa}$ at 28 days. High-strength concrete:
$60$--$100\,\text{MPa}$. Ultra-high-performance concrete (UHPC, with
silica-fume addition and steel-fibre reinforcement): $150$--$200\,
\text{MPa}$. Compressive strengths of $250\,\text{MPa}$ have been
achieved in research laboratories. Tensile strengths are
$10$--$15\,\%$ of compressive strengths; concrete is treated in
structural design as having essentially no tensile capacity, and the
tensile load is carried by reinforcing steel.

\engterm{Reinforced and prestressed concrete}. Steel reinforcement
bars (rebar) carry the tensile loads that concrete cannot. The
working bond between steel and concrete depends on bar surface
deformations (the lugs on a deformed bar), on the alkalinity of the
hydrating cement (which forms a protective passive layer on the steel),
and on the cover concrete that protects the steel from chloride
ingress (section~9 of this volume returns to the corrosion mechanism).
\engterm{Prestressed concrete} pre-loads the concrete in compression by
tensioning steel tendons before (pre-tensioning) or after (post-
tensioning) the concrete sets. The tendons, anchored at the ends,
transfer their tension to the concrete as compression that more than
offsets any tensile stresses the service load would otherwise produce.
Prestressing extends the working span of concrete structures from the
$\sim 10\,\text{m}$ practical for reinforced concrete to over $200\,
\text{m}$ for prestressed segmental box girders.

\engterm{Cement chemistry alternatives}. The carbon footprint of
ordinary Portland cement is approximately $0.8\,\text{t}$ of $\mathrm{CO_2}$
per tonne of cement, roughly half from calcination of limestone and
half from the kiln fuel. Cement production accounts for approximately
$8\,\%$ of global anthropogenic $\mathrm{CO_2}$ emissions, current as
of 2023. The chemistry alternatives include: replacement of clinker
with supplementary cementitious materials (ground granulated blast-
furnace slag, fly ash, calcined clays); calcium sulphoaluminate
cements; geopolymer cements (alkali-activated aluminosilicates); and
the magnesium-based reactive magnesia cements. Each trades
$\mathrm{CO_2}$ intensity for some combination of cost, availability,
strength, and durability; none is at production parity with Portland
cement as of 2025. Volume V Chapter 10 returns to this trade in the
materials-selection chapter.

\section{Electronic ceramics: piezoelectrics, ferroelectrics}\pathtag{standard}

A subset of ceramics has the crystal symmetry required to couple
mechanical and electrical fields. \engterm{Piezoelectric} ceramics
develop an electrical polarisation in response to applied mechanical
stress, and a mechanical strain in response to applied electrical
field. The two effects are governed by the constitutive equations
\begin{align*}
D_i &= d_{ij} \sigma_j + \epsilon^{\sigma}_{ik} E_k, \\
\varepsilon_j &= s^E_{jk} \sigma_k + d_{ij} E_i,
\end{align*}
where $D_i$ is the electric displacement, $\sigma_j$ the stress
(Voigt-notation), $E_k$ the electric field, $\varepsilon_j$ the
strain, $d_{ij}$ the piezoelectric charge constants, $\epsilon^\sigma$
the permittivity at constant stress, and $s^E$ the elastic compliance
at constant field. The working piezoelectric ceramics are
lead-zirconate-titanate (PZT, the workhorse of ultrasonic transducers,
fuel injectors, sonar, and piezo actuators), barium titanate (the
historic predecessor, still used for capacitors), lithium niobate (used
in single-crystal form for surface-acoustic-wave filters and
nonlinear-optical devices), and the lead-free alternatives (potassium-
sodium niobate, sodium-bismuth titanate) that are slowly displacing
PZT under European RoHS-style regulation.

\engterm{Ferroelectric ceramics} have a spontaneous polarisation that
can be switched by an applied electric field. PZT and barium titanate
are both ferroelectric below their Curie temperatures ($T_C \approx
360\,\degC$ for the most-used PZT compositions, $130\,\degC$ for
$\mathrm{BaTiO_3}$). The hysteresis loop $P(E)$ has a remanent
polarisation $P_r$ and a coercive field $E_c$, analogous to the
$M(H)$ hysteresis of a ferromagnet. The non-volatile memory cell of a
ferroelectric random-access memory (FeRAM) stores a bit in the
remanent polarisation direction.

\engterm{Working numbers for PZT}. Hard PZTs (intended for high-power
ultrasonic transducers) have piezoelectric charge constants $d_{33}
\approx 300\,\text{pC}/\text{N}$, coupling factors $k_{33} \approx
0.7$, and Curie temperatures $\sim 320\,\degC$. Soft PZTs (intended
for sensors and low-power actuators) have $d_{33} \approx 600$--$800\,
\text{pC}/\text{N}$, larger losses, and lower Curie temperatures.
Single-crystal relaxor ferroelectrics (PMN-PT, PZN-PT) reach $d_{33}
\sim 2000\,\text{pC}/\text{N}$ and are used in premium ultrasound
imaging probes.

\engterm{Dielectric and capacitor ceramics}. The bulk of the
capacitor market by part count is multilayer ceramic capacitors
(MLCCs) built from barium-titanate-based dielectrics. The relative
permittivity is engineered from $\sim 30$ (Class~I, temperature-
stable, low-loss) to $\sim 10{,}000$ (Class~II, ferroelectric, with
strong voltage and temperature dependence). The 2024 production of
MLCCs runs to several trillion units per year. Volume IV Chapter 9
develops the circuit-level use; this chapter notes only that the
ceramic chemistry is the constraint.

\engterm{Pyroelectric and electrocaloric ceramics}. A pyroelectric
develops a polarisation change with temperature, used in passive
infrared sensors (the motion-detection sensor in nearly every wall-
mounted occupancy switch). An electrocaloric ceramic develops a
temperature change with applied field, a candidate solid-state cooling
mechanism. Both effects share the polar-crystal-symmetry origin with
piezoelectricity.

\section{Failure: brittle fracture, thermal shock, concrete spalling}\pathtag{core}

A ceramic part fails in one of three working modes: brittle fracture
under monotonic or cyclic mechanical load, thermal shock under sudden
temperature change, or concrete spalling under combined chemical and
mechanical attack. Each has a quantitative criterion, a microstructural
mechanism, and a documented case in the engineering record.

\engterm{Brittle fracture under load}. The Griffith-Irwin analysis
gives the failure criterion: $K_I = Y \sigma \sqrt{\pi a} \geq K_{Ic}$.
The crack runs at the speed of an elastic wave (a few thousand metres
per second) once $K_I$ reaches $K_{Ic}$; there is no warning. The
fracture surface carries a fingerprint of the crack-initiation site (a
mirror region of smooth, slow growth surrounding the origin), a mist
region of slightly rougher growth, and a hackle region of fast
fracture and possible branching. The mirror-mist-hackle pattern is the
working diagnostic of brittle fracture and lets a fractographer trace
back to the originating flaw.

The Hubble Space Telescope's primary mirror is a brittle-ceramic part
in service \cite{acc:nasa-hubble-optical-systems-1990}. The mirror is
a ULE titania-silicate glass that survived launch loads, decades of
on-orbit thermal cycling, and the optical-correction installations of
the servicing missions. The 1990 spherical-aberration failure was an
optical figuring error, not a structural failure, but the structural
robustness of the mirror substrate against the launch environment was
itself a Weibull-statistics achievement: the substrate was proof-tested
to censor the lower tail of its strength distribution before flight.
The substrate's design strength reserve was set by the Weibull modulus
of the ULE glass and by the depth of the proof-test loading.

\engterm{Thermal shock}. A ceramic subjected to a temperature change
$\Delta T$ on a timescale short compared with the thermal-diffusion
time $\tau_{\text{diff}} = L^2/\alpha_{\text{th}}$ develops thermal
stresses
\[
\sigma_{\text{thermal}} \approx \frac{E \alpha \Delta T}{1 - \nu}.
\]
For a kiln furniture plate of alumina ($E = 380\,\text{GPa}$,
$\alpha = 8 \times 10^{-6}\,\si{\per\kelvin}$, $\nu = 0.23$) cooled
from $1200\,\degC$ to $200\,\degC$ in $30\,\text{s}$, the stress
amplitude on the surface is $\sigma \approx 3.3\,\text{GPa}$, an order
of magnitude above the working strength; the plate fractures into
multiple pieces. The two engineering responses are to lower
$\alpha$ (use cordierite or fused silica) and to lower the rate of
temperature change.

The same mechanism breaks the windshield of a car parked in winter
when the driver pours hot water on the frost-covered glass; it
cracks the porcelain dental crown that contacts a sudden draught of
ice water; it shatters the borosilicate beaker pulled from a Bunsen
burner and set on a cold bench. Each failure is the same equation
evaluated at a different scale.

\engterm{Concrete spalling}. Spalling is the loss of surface concrete
by mechanical failure of the cement-aggregate matrix. The working
mechanisms are several, all driven by an internal pressure that
exceeds the local tensile strength of the matrix:

\begin{itemize}[leftmargin=*]
\item \engterm{Freeze-thaw spalling}. Water in concrete pores expands
  by approximately $9\,\%$ on freezing. Concrete without air-
  entrainment admixtures cannot accommodate the expansion; the pore
  pressure rises until the matrix fails locally. Engineered air-
  entrainment introduces $4$--$8\,\%$ by volume of small bubbles that
  give the freezing water somewhere to go. Air-entrainment is now
  mandatory in concrete exposed to freeze-thaw cycling.
\item \engterm{Fire spalling}. Concrete exposed to fire above
  $\sim 300\,\degC$ loses water from its pores; in high-strength
  concretes ($w/c < 0.35$) the low permeability traps steam pressure,
  which spalls the surface off explosively. Tunnel-fire experience (the
  Mont Blanc tunnel fire of 1999, the Channel Tunnel fire of 1996)
  showed depths of spalling of $50$--$300\,\text{mm}$ on high-strength
  linings, and the engineering response has been to add polypropylene
  fibres that melt and provide pressure-release channels.
\item \engterm{Alkali-silica reaction (ASR) spalling}. Reactive silica
  aggregates (some andesites, opals, and cherts) react with the
  alkaline pore solution of concrete to form an expanding alkali-
  silicate gel. The expansion cracks and spalls the concrete from the
  inside, with damage developing over years to decades. The
  engineering responses are to test aggregates for reactivity, to use
  low-alkali cement, or to add lithium-based admixtures that suppress
  the reaction.
\item \engterm{Chloride-induced reinforcement corrosion}. Chloride
  ions from de-icing salt or seawater diffuse through concrete cover
  to the reinforcing steel, depassivating it and initiating corrosion.
  The corrosion product is $\sim 2$--$6$ times the volume of the
  consumed steel; the expansive pressure spalls the cover concrete
  off and exposes the reinforcement. Section~9 of this volume develops
  the corrosion mechanism; the failure mode is the canonical
  durability failure of reinforced concrete in marine or de-iced
  environments.
\end{itemize}

The four spalling mechanisms have a common pattern: an internal
expansion (water freezing, steam, alkali-silicate gel, corrosion
product) generates tensile stresses in the matrix that exceed its
brittle-fracture strength. The engineering response in each case is
the same in form: either eliminate the driving mechanism (low-alkali
cement, low-permeability concrete, ductile reinforcement, polypropylene
fibres), or accommodate the expansion (air-entrainment, drainage,
release channels).

\begin{principle}[A ceramic fails when an internal or external stress
exceeds the brittle-fracture strength of its weakest local element]
The strength is a random variable; the working stress must be set
below the lower tail of the strength distribution, not below its mean.
The engineering of structural ceramics is the engineering of the
lower-tail-of-strength distribution: by inspection, by surface
preparation, by prestress, by proof testing, by air-entrainment, by
fibre reinforcement, by acceptance of a higher inspection burden than
metals require. The failure section in any design review of a ceramic
part is the section that asks, "what is the probability that a flaw
larger than the inspection-detection limit exists, and what stress
will it carry?"
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Weibull characterisation of a brittle material]
The reader characterises the strength distribution of a brittle
material from a controlled bending or compression test on many
nominally identical specimens, and fits a Weibull distribution to the
data. The deliverable is a fifteen- to twenty-five-page report.

\textbf{Track}. Physical measurement plus analysis.
\textbf{Hazard class}: low (chalk dust; safety glasses for sample
handling).

\textbf{Materials}.

\begin{itemize}[leftmargin=*]
\item A box of standard chalkboard chalk sticks (calcium sulphate or
  calcium carbonate, depending on the brand), or cast plaster bars,
  or mortar bars. Thirty specimens minimum, fifty preferred.
\item A simple three-point bending fixture: two support pins
  $20\,\text{mm}$ apart on a stiff base, a loading rod, and a method
  of applying load (a kitchen scale with a screw-down jig, or a
  laboratory force gauge if available).
\item Calipers, a notebook, a camera.
\end{itemize}

\textbf{Procedure}.

\begin{itemize}[leftmargin=*]
\item Measure the cross-section and length of every specimen to
  $\pm 0.1\,\text{mm}$; record.
\item Mount each specimen on the support pins and load to fracture in
  three-point bending. Record the peak force. Photograph the fracture
  surface of every fifth specimen for later fractographic inspection.
\item Convert peak force to bending stress: $\sigma_f = 3 F L / (2 b
  d^2)$ for a rectangular cross-section ($b$ width, $d$ depth, $L$
  span), or $\sigma_f = 8 F L / (\pi D^3)$ for a circular cross-
  section of diameter $D$.
\end{itemize}

\textbf{Analysis}.

\begin{itemize}[leftmargin=*]
\item Sort the $N$ strength values in ascending order and compute the
  empirical cumulative probability of failure $P_{f,i} = (i - 0.5)/N$.
\item Plot $\ln \ln[1/(1 - P_f)]$ versus $\ln \sigma$ and fit a
  straight line by least squares. Report the Weibull modulus $m$
  (the slope) and the scale parameter $\sigma_0$ (from the intercept).
\item Compute the $90\,\%$ confidence interval on $m$ by a bootstrap
  resampling procedure (resample $N$ strength values with replacement
  $10^3$ times; refit $m$ each time; report the $5$th and $95$th
  percentiles).
\item Inspect the fractographs of the failed specimens. Identify the
  apparent origin (mirror centre) on at least three specimens and
  measure its position relative to the tensile surface. Discuss
  whether the largest-flaw assumption is supported.
\end{itemize}

\textbf{Reflection ($1{,}500$ to $2{,}500$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
\item What is $m$ for the chosen material? How does the value compare
  with the working ranges for engineering ceramics?
\item What is the predicted strength at a survival probability of
  $99.9\,\%$ for a specimen of the test volume? At a specimen of
  $10\times$ the test volume?
\item What sources of scatter are visible in the data, and which are
  artefacts of the test (misalignment, span variability, surface
  damage during handling) rather than properties of the material?
\item One feature of the data that the reader had not anticipated.
\end{itemize}

\textbf{Deliverable}. The raw data log, the analysis script (Python,
R, or spreadsheet), the report, and a brief reflection on the
limitations of bench-scale testing for a material whose engineering
use depends on volume scaling.

\textbf{Time}. Four to six hours of testing, four to eight hours of
data reduction and analysis, four to six hours of write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A soda-lime glass plate has an elastic modulus $E = 70\,\text{GPa}$,
fracture surface energy $\gamma_s = 0.3\,\text{J}/\text{m}^2$, and a
surface scratch $30\,\si{\micro\meter}$ deep. Estimate the tensile
stress at which the plate will fracture, using the Griffith criterion.
\end{exercise}

\begin{exercise}
An alumina specimen of volume $1\,\text{cm}^3$ has a measured Weibull
scale parameter $\sigma_0 = 350\,\text{MPa}$ at a failure probability
of $63\,\%$, and a Weibull modulus $m = 10$. Estimate the working
strength at the same failure probability for a specimen of volume
$100\,\text{cm}^3$.
\end{exercise}

\begin{exercise}
A silicon-carbide turbine vane has a working stress of $250\,\text{MPa}$,
a Weibull scale parameter $\sigma_0 = 500\,\text{MPa}$ at a reference
volume of $V_0 = 10\,\text{cm}^3$, and a Weibull modulus $m = 15$. The
working volume of the vane is $V = 50\,\text{cm}^3$. Compute the
probability of failure per loading cycle.
\end{exercise}

\begin{exercise}
A piezoelectric stack actuator is built from PZT plates with $d_{33} =
500\,\text{pC}/\text{N}$, area $A = 100\,\text{mm}^2$, thickness
$t = 0.5\,\text{mm}$, and $n = 100$ plates connected mechanically in
series and electrically in parallel. Compute the free displacement at
an applied voltage of $200\,\text{V}$ and the blocking force at the
same voltage. The Young's modulus of the PZT is $50\,\text{GPa}$.
\end{exercise}

\begin{exercise}
A reinforced-concrete beam carries a service moment that produces a
peak tensile stress of $5\,\text{MPa}$ in the concrete extreme fibre.
Given a 28-day compressive strength of $40\,\text{MPa}$ and an
estimated tensile strength of $10\,\%$ of the compressive value,
assess the safety against cracking and explain why this is not the
governing design check.
\end{exercise}

\begin{exercise}
A tempered glass window has a surface residual compression of
$-100\,\text{MPa}$ and an annealed-glass surface strength (at the
same surface-flaw distribution) of $40\,\text{MPa}$. Compute the
effective tensile strength of the tempered window.
\end{exercise}

\begin{exercise}
A pyroceramic stovetop element has a thermal expansion coefficient
$\alpha = 0.5 \times 10^{-6}\,\si{\per\kelvin}$, an elastic modulus
$E = 90\,\text{GPa}$, and a Poisson ratio $\nu = 0.25$. Compute the
thermal stress induced by a $\Delta T = 400\,\si{\kelvin}$ across the
thickness of the element. Compare with a typical fracture strength of
$200\,\text{MPa}$.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Starting from the elastic strain energy in a cracked plate and the
surface energy associated with the new surfaces, derive the Griffith
criterion $\sigma_f = \sqrt{2 E \gamma_s / (\pi a)}$ for plane stress.
\end{exercise}

\begin{exercise}
Derive the linearised plotting form of the Weibull distribution,
$\ln \ln[1/(1 - P_f)] = m \ln \sigma + \text{const}$, and identify
the slope and intercept with the Weibull parameters $m$ and
$\sigma_0$.
\end{exercise}

\begin{exercise}
Derive the volume-scaling relation $\sigma_1/\sigma_2 = (V_2/V_1)^{1/m}$
from the Weibull distribution and the assumption that the flaws are
distributed uniformly in volume.
\end{exercise}

\begin{exercise}
Show that the radius ratio for an octahedrally coordinated cation is
$r^+/r^- \geq 0.414$ by a geometric argument on a hard-sphere model.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the energy required to produce $1\,\text{kg}$ of Portland
cement, using the heat of calcination of limestone and an assumed
kiln efficiency.
\end{exercise}

\begin{exercise}
Estimate the time required for $50\,\text{mm}$ of chloride to diffuse
through a concrete cover from a seawater splash zone, assuming an
effective diffusion coefficient $D = 10^{-12}\,\text{m}^2/\text{s}$.
\end{exercise}

\begin{exercise}
Estimate the maximum useful piezoelectric voltage developed by a PZT
disc of diameter $20\,\text{mm}$ and thickness $1\,\text{mm}$ struck by
a steel ball that delivers a $1\,\text{N}$ impulse over $1\,\text{ms}$.
\end{exercise}

\begin{exercise}
Estimate the freeze-thaw expansion pressure developed in a saturated
concrete pore when the water turns to ice, given the $9\,\%$ volume
expansion of ice and the bulk modulus of water at near-freezing
conditions.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a Monte Carlo simulation of the strength of a brittle bar
under bending: place $N_f$ flaws of random length (drawn from an
exponential distribution with mean $\bar{a}$) at random positions, and
compute the bar's fracture stress as the minimum applied stress at
which any flaw becomes critical under the Griffith criterion. Run
$10^4$ replicate bars and fit a Weibull distribution to the simulated
strengths. Compare the fitted Weibull modulus with the predicted value
from the flaw-length distribution.
\end{exercise}

\begin{exercise}
Implement a one-dimensional finite-difference solver for transient
heat conduction in a ceramic plate of thickness $L$, subjected to a
step change in surface temperature. Plot the temperature profile and
the corresponding thermal stress profile as a function of time. Identify
the time at which the peak thermal stress is reached.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify a ceramic substrate for an automotive catalytic-converter
monolith that must survive $10^4$ thermal cycles between $20\,\degC$
and $900\,\degC$ over a $200{,}000\,\text{km}$ design life. State the
required thermal-expansion coefficient, the maximum permissible
expansion mismatch with the wash-coat, and the rationale for choosing
between cordierite and silicon carbide.
\end{exercise}

\begin{exercise}
Design a chemically strengthened glass cover for a smartphone display
that must survive a drop test from $1.5\,\text{m}$ onto a hard floor.
State the required compressive depth, the surface compression value,
and the depth of compression chosen.
\end{exercise}

\begin{exercise}
Specify the mix design for a $50\,\text{MPa}$ concrete intended for a
multi-storey building's structural columns in a freeze-thaw climate.
State the $w/c$ ratio, the cement content, the air-entrainment
fraction, and the curing regime.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A batch of cast-alumina kiln-furniture plates has fractured in service
at a rate of $5\,\%$ per heating cycle. Inspection of fracture surfaces
shows a consistent mirror region at the tensile surface near the centre
of the plate. Diagnose the failure mode and propose a single change to
the firing or handling procedure that could halve the failure rate.
\end{exercise}

\begin{exercise}
A 30-year-old reinforced-concrete bridge deck exhibits parallel cracks
along the line of the reinforcing bars and rust staining at the cracks.
Diagnose the failure mode, propose an intervention to arrest its
progression, and estimate the remaining service life assuming the
intervention is successful.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A glass cooktop manufactured from lithium-aluminosilicate
glass-ceramic shatters when a pot of cold water is placed on a hot
burner. Identify the dominant failure mechanism, the role (if any) of
the near-zero expansion coefficient, and the manufacturing or design
factor that must have failed.
\end{exercise}

\begin{exercise}
A high-strength-concrete column in an underground car park spalls
violently during a vehicle fire. Identify the failure mode, the
critical concrete-property parameter, and the engineering remediation
that production high-strength concrete now incorporates.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A colleague proposes to specify a ceramic part using only the
mean and standard deviation of strength measured on a small coupon
batch, without fitting a Weibull distribution. Write a one-paragraph
reply explaining why the mean-and-standard-deviation framing is
inadequate and what the consequence is for the working safety factor.
\end{exercise}

\begin{exercise}
A start-up's pitch proposes a high-strength glass that achieves its
strength by exceeding the surface-flaw-handling capacity of the
production line. Identify the design-versus-production gap, the
likely manufacturing failure mode, and the question to ask in the next
investor meeting.
\end{exercise}
