\chapter{Manufacturing-driven materials properties}
\label{vol05:ch11}

\epigraph{The properties of a finished engineering part are the
properties of the alloy modified by the manufacturing route the part
travelled. A wrought aluminium plate, a cast aluminium engine block,
and a selectively-laser-melted aluminium aerospace bracket made of
the same alloy have different microstructures, different residual
stresses, different anisotropies, and different failure modes.}{Adapted
from S.\ Kalpakjian and S.\ Schmid, working framing of manufacturing
processes \cite[ch.~1, pp.~1--12]{text:kalpakjian-manufacturing}.}

\chapmeta{Half-life: durable. Archetypes: scaling, failure. Exercise target: 31. Project track: physical measurement of additively manufactured samples. Hazard class: medium (printed parts, surface treatments, lifting and machining; safety glasses, hearing protection, hot-surface awareness). Reader path default: standard, with sections explicitly tagged. Named cases: Hyatt Regency walkway fabrication-modification (1981); Hubble primary mirror polishing (1990); welds and forging discontinuities described by mechanism only for in-service components.}

\noindent
A part's working properties are set as much by its manufacturing
history as by its alloy chemistry. The chapter develops this thesis on
engineering footing: the residual stresses from casting, forging, and
rolling; the heat-affected zone of welded joints; the surface
treatments (case hardening, shot peening, anodising) that engineer the
near-surface mechanical state; the anisotropy, porosity, and residual
stress characteristic of additively manufactured parts; the
non-destructive-testing methods used to certify them; and the failure
mode that recurs when the gap between as-designed and as-manufactured
properties is not accounted for in the design. The chapter is the
discipline-side counterpart of Volume V Chapter 10: selection chooses
the material; processing determines what the material actually
becomes.

\section{Why how it was made matters as much as what it is}\pathtag{core}

A material in service is a microstructure: a specific arrangement of
grains, phases, dislocations, voids, and residual stresses. The
microstructure is set by the processing history. The same nominal
alloy, processed by different routes, gives different microstructures
and therefore different properties.

\engterm{The same alloy, three routes}. Aluminium alloy 7075 in the
T6 condition is a specification: 5.6\% Zn, 2.5\% Mg, 1.6\% Cu, 0.23\%
Cr, balance Al, solution-treated and aged to peak strength. The
specification fixes the chemistry. The microstructure depends on the
processing route:

\begin{itemize}[leftmargin=*]
\item \engterm{Wrought 7075-T6 plate}. Hot-rolled at $\sim 380$ -- $420\,
  \degC$ from a cast ingot, then cold-rolled to final thickness,
  solution-treated at $\sim 470\,\degC$, water-quenched, and aged at
  $\sim 120\,\degC$ for $24\,\text{h}$. Microstructure: pancaked
  grains elongated in the rolling direction, fine and uniform
  $\eta'$ precipitates, low porosity, in-plane anisotropy of strength
  and toughness ($\sim 10$-$20\,\%$ between L and T directions in
  yield strength and $\sim 30\,\%$ in $K_{Ic}$). Working yield
  strength: $\sim 500\,\text{MPa}$.
\item \engterm{Sand-cast 7075}. Poured into a sand mould at $\sim
  700\,\degC$, solidified over minutes, then solution-treated and
  aged. Microstructure: equiaxed grains $\sim 100$ -- $500\,\si{\micro\meter}$
  in size, interdendritic porosity ($\sim 1$ -- $3\,\%$), coarse
  primary phases that do not fully dissolve in solution treatment.
  Working yield strength: $\sim 350$ -- $400\,\text{MPa}$.
\item \engterm{Laser powder-bed fusion (L-PBF) 7075}. Built layer by
  layer from atomised powder at $\sim 30\,\si{\micro\meter}$ layer
  thickness, with each layer melted by a focused laser at $\sim 200\,
  \text{W}$ traversing at $\sim 1\,\text{m}/\text{s}$. Microstructure:
  fine columnar grains aligned with the build direction, residual
  stresses up to $\sim \sigma_y$ in magnitude, build-direction
  anisotropy ($\sim 10$ -- $20\,\%$ between vertical and horizontal
  build orientations), porosity $\sim 0.1$ -- $1\,\%$ depending on
  process control. Working yield strength: $\sim 400$ -- $500\,
  \text{MPa}$ as-built; $\sim 500$ -- $600\,\text{MPa}$ after stress-
  relief and HIP. The 7075 alloy is in fact notoriously difficult to
  print without solidification cracking; production aerospace L-PBF
  uses related alloys (AlSi10Mg, Scalmalloy) tailored for the process.
\end{itemize}

The three routes share the chemistry; they do not share the working
property. A design verified at the wrought-plate property level cannot
be substituted with a cast or L-PBF part of the ``same'' alloy
without re-qualification.

\engterm{Process-driven property variation in metals}. The dominant
mechanisms by which manufacturing alters working properties:

\begin{itemize}[leftmargin=*]
\item \engterm{Grain size}. The Hall-Petch relation $\sigma_y =
  \sigma_0 + k_y / \sqrt{d}$ gives yield strength as a function of
  grain size $d$. A factor of $100$ in grain size (millimetres in a
  large casting, micrometres in a fine-grain L-PBF deposit) gives a
  factor of $10$ in $\sqrt{d}$ and a corresponding scaling in
  Hall-Petch yield-strength contribution.
\item \engterm{Phase volume fractions and morphologies}. Hot-worked
  steel has different carbide distributions from as-cast steel.
  Solution-treated and aged precipitation hardening alloys have
  different precipitate sizes after different ageing schedules.
\item \engterm{Residual stresses}. Self-equilibrated stresses in a
  part with no external load. Forging, casting, rolling, welding,
  surface treatments, and additive manufacturing all leave residual
  stresses. The working magnitude is $0.1$ -- $1.0$ times the yield
  strength.
\item \engterm{Defects}. Porosity (cast and additively manufactured
  parts), inclusions (cast and wrought parts), lack-of-fusion defects
  (welds, AM), folds and laps (forgings), seams (rolled bars).
\item \engterm{Crystallographic texture}. Preferred crystal
  orientations developed by deformation processing. A heavily rolled
  steel sheet has a texture (e.g.\ rolling-direction $\{100\}\langle011\rangle$
  cube texture) that gives anisotropic properties: different yield
  strength in L, T, and $45^\circ$ directions; different formability
  (the Lankford coefficient $r$ varies with direction).
\item \engterm{Surface state}. Surface roughness, residual stress
  near the surface, work-hardened or chemically-modified surface
  layers (carburised, nitrided, anodised, coated).
\end{itemize}

\begin{principle}[The working property of a manufactured part is the
working property of the manufactured microstructure, not of the
nominal alloy]
A part qualified to a property requirement is qualified at the
specific combination of alloy and process. A part of the same alloy
made by a different process is a different engineering part. The
design discipline names the manufacturing route in the same
specification line as the alloy; the qualification discipline tests
the route's output, not the alloy's nominal properties.
\end{principle}

\section{Forging, casting, rolling: residual stresses}\pathtag{standard}

The three primary metal-forming routes leave characteristic residual
stress patterns that the design must accommodate.

\engterm{Casting and the resulting residual stresses}. A liquid metal
poured into a mould solidifies from the mould walls inward. The
solidifying shell cools and contracts; the interior, still liquid,
adjusts. By the time the interior is solid, the shell has contracted
and the interior is now contracting against the constraint of the
shell. The result, after full cooling: the shell is in tension, the
interior is in compression. The magnitude depends on the thermal-
contraction coefficient, the temperature drop, and the elastic
modulus of the alloy: $\sigma_{\text{res}} \approx E \alpha \Delta T
/ (1 - \nu)$ in the elastic estimate. For aluminium cast and cooled
through $400\,\si{\kelvin}$, $\sigma_{\text{res}} \approx 70 \times
22 \times 10^{-6} \times 400 / 0.77 \approx 800\,\text{MPa}$ in the
elastic estimate, well above the yield strength: the alloy yields
during cooling and the residual stress is limited by the alloy's
yield strength at the contraction temperature. Net residual stresses
in as-cast aluminium are typically $\sim 100$ -- $200\,\text{MPa}$ in
the shell, balanced by compressive residual stresses in the interior
\cite[ch.~10]{text:dieter-mechanical-metallurgy}.

\engterm{Working remediations for casting residual stress}.
Stress-relief annealing at $200$ -- $400\,\degC$ for aluminium, $550$
-- $650\,\degC$ for steel: the elevated temperature reduces the alloy's
yield strength, so the residual stress relaxes plastically; on cooling
from the stress-relief temperature, lower thermal gradients leave
smaller residual stresses. Hot isostatic pressing (HIP) at temperatures
and pressures that close internal porosity and homogenise the
microstructure.

\engterm{Forging residual stresses}. Forging deforms the metal at
elevated temperature, typically just below the recrystallisation
temperature for ``warm forging'' or above it for ``hot forging.''
The deformation is non-uniform: parts of the workpiece deform more
than others; the cooler die surface restrains the workpiece surface,
which deforms less than the interior. The result, after cooling, is
a residual stress pattern set by the non-uniform deformation. Cold
forging adds plastic-deformation residual stresses; hot forging
followed by rapid cooling adds thermal residual stresses on top.

The engineering response: a normalising or annealing heat treatment
after forging redistributes the residual stresses, the design
explicitly accommodates the working magnitude (typically $20$ -- $40\,\%$
of $\sigma_y$ in well-controlled forgings), and the final machining
of critical surfaces is done after the heat treatment so that surface
machining stresses do not stack on top.

\engterm{Rolling residual stresses}. Hot rolling at $\sim 0.6$ -- $0.8\,
T_m$ leaves modest residual stresses because the metal is in the
elastic-plastic regime where plastic strains relax thermal stresses.
Cold rolling leaves substantial residual stresses with a
characteristic through-thickness pattern: the surface is in
compression (if the rolls bit into the surface and dragged metal
through), the core in tension. The magnitudes are $0.3$ -- $0.7\,\sigma_y$
in the surface for heavy reductions. Cold-rolled sheet that is not
stress-relieved develops a tendency to spring back, to curl, and to
distort during subsequent forming.

\engterm{Quenching residual stresses}. Quenching from solution-
treatment temperature creates the most severe thermal residual
stresses encountered in normal manufacturing. The surface cools
rapidly, contracts, and reaches a low temperature while the interior
is still hot and contracting; the surface is plastically deformed by
the contraction of the interior. As the interior continues to cool
and contract, it pulls against the now-cooled surface; the final
state has the surface in compression and the interior in tension. The
working magnitudes are $0.5$ -- $1.0\,\sigma_y$, and the surface
compression is engineering useful: a quenched-and-tempered steel
bar has compressive surface residual stresses that improve fatigue
performance.

The Almen strip provides a working measurement of surface compressive
residual stress for shot peening (section~11.4).

\engterm{Computational residual-stress prediction}. Finite-element
codes for thermomechanical analysis of casting and welding (SYSWELD,
ABAQUS welding modules, in-house codes for power-generation OEMs)
predict residual stresses for representative parts to within $\sim 30\,\%$
of measured values, which is adequate for design-by-analysis. The
calibration to specific alloys and process parameters is the working
limit of the predictive accuracy. The validation by neutron-diffraction
or contour-method measurement is the working closure.

\section{Heat-affected zone in welding}\pathtag{core}

Welding joins two metal parts by local melting and resolidification.
The metal adjacent to the molten weld pool reaches temperatures that
modify its microstructure without melting it. The \engterm{heat-affected
zone (HAZ)} is the working region of altered properties.

\engterm{The temperature-profile-driven HAZ structure}. The HAZ
of a steel weld, on a cross-section, shows distinct subzones from
the fusion line outward:

\begin{itemize}[leftmargin=*]
\item \engterm{Coarse-grained HAZ}: metal that reached $\sim 1100$ -- $1400\,
  \degC$. Austenite grain size grows; on cooling, the coarse-grain
  austenite transforms to coarse-grain ferrite, bainite, or martensite
  depending on cooling rate. Coarse-grain martensite is high in
  hardness, low in toughness, and the canonical source of HAZ
  cracking in carbon and low-alloy steels.
\item \engterm{Fine-grained HAZ}: metal that reached $\sim 900$ -- $1100\,
  \degC$. Austenite grain growth is limited; on cooling, the structure
  is fine-grain ferrite-pearlite or fine bainite. Properties are good.
\item \engterm{Intercritical HAZ}: metal that reached $A_{c1} < T <
  A_{c3}$ (the inter-critical region between the two phase-transformation
  temperatures of the steel). Partial austenitisation; on cooling,
  a mixed structure of new (low-carbon) ferrite and high-carbon
  islands. Toughness is generally adequate but local hard spots can
  initiate cracks.
\item \engterm{Subcritical HAZ}: metal that reached $T < A_{c1}$.
  No austenitisation. The metal experiences thermal tempering;
  pre-existing martensite is tempered, pre-existing ferrite-pearlite
  is little changed. The subcritical zone is the broadest part of
  the HAZ in thick sections.
\end{itemize}

The HAZ is wide in slow welds (manual stick welding, oxy-fuel cutting)
and narrow in fast welds (laser welding, electron-beam welding, friction
stir welding). Working HAZ widths: $\sim 10$ -- $30\,\text{mm}$ for
SMAW on $25\,\text{mm}$ plate; $\sim 1$ -- $3\,\text{mm}$ for laser
welding on the same plate.

\engterm{Hardenability and hydrogen cracking}. Steels with high
hardenability (high carbon, high alloy content, large section size)
form martensite in the coarse-grain HAZ on cooling, which is hard and
crack-susceptible. Hydrogen introduced during welding (from moisture
in the flux, from cellulose-coated electrodes, from contaminated weld
joints) is trapped at martensitic regions and produces hydrogen-
assisted (cold) cracking at temperatures below $200\,\degC$, often
hours to days after the weld is completed (delayed cracking). The
working prevention: preheat (raise the metal temperature so cooling
is slower and martensite formation is suppressed or tempered);
low-hydrogen consumables (basic-flux SMAW electrodes, low-moisture
storage, dry shielding gas); post-weld heat treatment (tempers any
martensite that did form).

\engterm{Sensitisation of austenitic stainless steel}. The HAZ of an
austenitic stainless-steel weld can be sensitised in the $500$ -- $800\,
\degC$ temperature range, where chromium-carbide precipitation at
grain boundaries depletes adjacent matrix of chromium and produces
intergranular-corrosion susceptibility (Volume V Chapter 9 develops
the corrosion side). The working remediation: low-carbon grades
(304L, 316L, with $C < 0.03\,\%$) or stabilised grades (321 with Ti,
347 with Nb) that preferentially form titanium or niobium carbides
in solution and prevent chromium-carbide precipitation in the HAZ.

\engterm{Reheat cracking in nickel-base alloys and creep-resistant
steels}. Service in the creep range (or post-weld heat treatment at
elevated temperature) can produce intergranular cracks in the
coarse-grained HAZ of nickel-base superalloys and certain creep-
resistant steels. The mechanism involves precipitate-pinned grain
boundaries that fail to relax stress concentrations at triple points.
The working response is alloy choice (avoiding susceptible grades
where possible), filler metal choice (avoiding overmatched-strength
filler), and post-weld heat-treatment schedules tuned to specific
alloy-and-section combinations.

\engterm{Weld residual stresses}. A weld is the canonical thermal-
cycle source of residual stresses. The fusion zone solidifies in
contact with the surrounding cold parent metal; as the fusion zone
cools, it contracts against the constraint of the parent metal; the
final residual stress in the fusion zone and the adjacent HAZ is
tensile, of magnitude approaching the yield strength of the weld
metal at the parent-metal temperature. The transverse residual stress
is smaller but still substantial.

The working implication: a structural weld carries a yield-magnitude
tensile residual stress in service that adds to the applied service
stress. The design fatigue life calculation must include the residual
stress as a mean-stress shift (Goodman correction). The standard
remediations are post-weld heat treatment (stress-relief at $\sim
600\,\degC$ for steel reduces residual stress to $\sim 0.2$ -- $0.3\,
\sigma_y$) and surface peening (introduces compressive residual stress
at the weld toe, the most fatigue-critical location).

\engterm{Working welding processes by HAZ profile}. SMAW (manual stick
welding): wide HAZ, moderate heat input, oldest and cheapest, the
hand-trade workhorse. GMAW/MIG (continuous wire, inert gas): narrower
HAZ, semi-automated, the modern fabrication-shop workhorse. GTAW/TIG
(tungsten electrode, hand-fed filler, inert gas): narrowest of the
arc processes, the precision-aerospace process. Laser welding: very
narrow HAZ, very fast, but joint preparation and clamping demanding.
Electron-beam welding: narrow HAZ, fast, but the workpiece must be
in vacuum. Friction stir welding: solid-state process at $\sim
0.8\,T_m$ peak temperature; no melt, much narrower HAZ, weld
properties often $80$ -- $90\,\%$ of parent metal versus $\sim 60\,\%$
for fusion welds in aluminium; the working process for aerospace
aluminium structure (the Space Shuttle external tank, Eclipse 500
aircraft fuselage, the C-Bus rail-car body).

\section{Surface treatments: case hardening, shot peening, anodising}\pathtag{standard}

The near-surface mechanical state of a part is the working region for
fatigue, wear, and corrosion. Surface treatments engineer this region
without changing the bulk.

\engterm{Case hardening of steel}. Carburising (gas, salt-bath, or
pack), nitriding (gas, plasma), carbonitriding, and surface induction-
or laser-hardening produce a high-hardness ``case'' over a tough
core. Working depths and hardnesses:

\begin{itemize}[leftmargin=*]
\item \engterm{Gas carburising}: case depth $0.5$ -- $2.5\,\text{mm}$,
  surface hardness $\sim 60\,\text{HRC}$, achieved by holding the
  part at $\sim 900\,\degC$ in a carbon-rich atmosphere for hours,
  then quenching. The canonical gear-tooth and bearing-race treatment.
\item \engterm{Nitriding}: case depth $0.1$ -- $0.5\,\text{mm}$,
  surface hardness $\sim 65$ -- $70\,\text{HRC}$, achieved by holding
  the part at $\sim 500$ -- $570\,\degC$ in ammonia or in plasma for
  $20$ -- $100\,\text{h}$. The case is hard, the diffusion is slow, the
  resulting compressive residual stresses are substantial. The
  canonical treatment for cam lobes, crankshafts, and high-precision
  bearing surfaces.
\item \engterm{Induction hardening}: a high-frequency induction coil
  heats a thin surface layer ($1$ -- $5\,\text{mm}$ depth, depending
  on frequency: $\sim 5$ -- $50\,\text{kHz}$ for medium-depth,
  $\sim 100$ -- $400\,\text{kHz}$ for thin); the heated surface is
  immediately quenched. The process is fast, controllable, and
  applies to medium-carbon steels (0.30 -- 0.55\% C). The canonical
  treatment for camshaft lobes, axle journals, and gear teeth where
  carburising is not used.
\end{itemize}

\engterm{Shot peening}. A controlled stream of small steel or
ceramic shot, typically $0.5$ -- $1\,\text{mm}$ diameter, is fired
at the part surface at velocities of $50$ -- $80\,\text{m}/\text{s}$.
Each impact plastically deforms a small surface dimple; the
surrounding elastic material constrains the dimple; the net effect is
a compressive residual stress in the near-surface layer, peaking at
$\sim 0.5$ -- $0.7\,\sigma_y$ at a depth of $\sim 0.1$ -- $0.3\,\text{mm}$.

The compressive residual stress shifts the mean stress on any
near-surface fatigue crack into compression; the working fatigue life
of shot-peened parts is typically $2$ -- $10\,\times$ longer than
unpeened parts at the same applied load. Working applications:
helicopter rotor pitch links, aircraft landing-gear components,
high-strength fasteners, automotive coil springs, gear teeth.

The Almen strip is the working calibration of shot-peening intensity:
a thin steel strip, peened on one side, curls because the peening
introduces compressive stress in the peened side; the curvature, the
Almen intensity, is the calibrated process variable.

\engterm{Anodising of aluminium}. An electrolytic process that grows
the aluminium oxide surface film from $\sim 20\,\text{nm}$ (natural
passive film) to $5$ -- $25\,\si{\micro\meter}$ (decorative anodising)
or $25$ -- $100\,\si{\micro\meter}$ (hardcoat anodising). The
thick oxide is hard ($\sim 1000\,\text{HV}$), corrosion-resistant
(the working aerospace and architectural finish for aluminium), and
electrically insulating. The oxide is brittle and porous; the pores
are typically sealed by hot-water immersion or by impregnation with
dye and then sealant. The working trade-off: the anodised layer
slightly degrades fatigue life of the substrate because it acts as
a stress-concentrating brittle surface; fatigue-critical aluminium
parts are designed with shot peening before anodising to compensate.

\engterm{Plating}. Electroplated coatings (chromium for hard surfaces;
nickel and zinc for corrosion protection; cadmium for aerospace
fasteners; gold for electrical contacts; copper as a base layer)
deposit a metal layer of controlled thickness on a conducting
substrate. The plating layer has a characteristic microstructure
(typically columnar grains aligned with the plating direction) and
characteristic residual stresses (usually tensile, often substantial:
$300$ -- $1000\,\text{MPa}$ for hard chrome). Hard chromium plating on
aircraft landing-gear axles is a working aerospace example of a
tensile-residual-stress plating; the working remediation is to plate
slightly thicker than required and then shot-peen the assembled
plating to redistribute the residual stress.

\engterm{Conversion coatings, paint primers, and the corrosion-
protection stack}. The full surface stack for an aerospace aluminium
structure typically includes: alkaline clean, anodise (chromic acid
or sulfuric acid), conversion coating (chromate or trivalent chromium
replacement), primer (epoxy or polyurethane), topcoat (polyurethane,
fluoropolymer). Each layer has a function: clean removes contaminants,
anodise provides the corrosion-resistant base, conversion coating
improves paint adhesion, primer carries pigments and corrosion
inhibitors, topcoat resists UV and weather. The stack is engineered
as a system; substituting a single layer typically requires
re-qualification of the system.

\section{Additive manufacturing: anisotropy, porosity, residual stress}\pathtag{standard}

Additive manufacturing (AM) builds a part by depositing material in
controlled patterns, layer by layer. The seven ASTM-standard process
families \cite[ch.~3]{text:gibson-rosen-stucker-am} are:

\begin{itemize}[leftmargin=*]
\item \engterm{Powder bed fusion} (PBF): selective laser melting of
  metal powder (L-PBF: laser powder bed fusion), electron-beam
  melting (EBM: a fast process for refractory alloys), and selective
  laser sintering (SLS: polymer powders).
\item \engterm{Directed energy deposition} (DED): metal powder or
  wire fed into a laser-, electron-, or arc-generated melt pool;
  enables on-existing-part deposition, large-part build, and metal-
  metal repair.
\item \engterm{Vat photopolymerisation}: photopolymer resin cured by
  UV or LED light, layer by layer (SLA, DLP, CLIP).
\item \engterm{Material extrusion} (FDM/FFF): thermoplastic filament
  melted and extruded layer by layer.
\item \engterm{Material jetting}: photopolymer or wax droplets jetted
  and cured.
\item \engterm{Binder jetting}: binder droplets selectively bind
  powder; the bound part is then sintered (metal, ceramic) or
  infiltrated (sandstone, plaster).
\item \engterm{Sheet lamination}: sheets of paper, polymer, or
  metal bonded layer by layer.
\end{itemize}

\engterm{Process-driven property anisotropy in AM metals}. L-PBF and
EBM build parts layer by layer; the columnar grain structure aligns
with the build direction; the in-plane (XY) and build-direction (Z)
properties differ. Working anisotropy ratios: $\sigma_y$ in Z is
typically $\sim 90\,\%$ of $\sigma_y$ in XY; ductility in Z is
typically $\sim 50$ -- $80\,\%$ of XY; fatigue properties in Z are
typically the worst-direction, reflecting both the columnar grain
structure and the layerwise defect distribution (defects align with
layer boundaries).

\engterm{Porosity}. AM parts have characteristic defect populations
that the bulk-cast or wrought alternative lacks. Common defect
types:

\begin{itemize}[leftmargin=*]
\item \engterm{Gas porosity}: spherical or near-spherical pores from
  trapped gas (argon, nitrogen) in the powder or from gas evolved
  during melting. Typical sizes $10$ -- $100\,\si{\micro\meter}$,
  typical volume fractions $0.01$ -- $0.5\,\%$.
\item \engterm{Lack-of-fusion defects}: irregular, often
  through-layer voids where the laser energy was insufficient to
  fully melt the powder. The voids are typically $50$ -- $500\,
  \si{\micro\meter}$ and form sharp-tipped cracks aligned with the
  layer boundaries. These are the most fatigue-critical AM defect.
\item \engterm{Keyhole porosity}: deep, gas-filled pores in the
  centre of the melt pool when the laser energy is too high and the
  vapour pressure of the melt drills a keyhole that collapses on
  itself.
\item \engterm{Inclusions}: oxide skins on the powder particles
  carried into the melt pool, particularly for reactive alloys
  (aluminium, titanium) processed in argon with trace oxygen.
\end{itemize}

\engterm{Process control for AM defects}. The working process
parameters that bind defect formation: laser power, scan speed,
hatch spacing, layer thickness, beam-focus offset, scan-pattern
strategy, and atmosphere composition. The working window for a
specific alloy on a specific machine is typically only a few
percent wide on each parameter; deviation produces porosity at the
high or low ends, lack-of-fusion at the low end, and keyholing at
the high end. The qualification of an AM process is the
demonstration of the working window and the in-process monitoring
that confirms operation within it.

\engterm{Residual stresses in AM}. The layer-by-layer cooling cycle
creates large residual stresses in AM parts. Each new layer
solidifies on top of the cooled previous layer; the new layer
contracts on cooling and is constrained by the substrate below;
the net residual stress in the new layer is tensile in the build
plane, balanced by compressive stresses below. The cumulative effect
over many layers gives a part with characteristic stress patterns
(tensile near the top surface, compressive in the interior, tensile
at the base).

The magnitudes are large: residual stresses approaching the yield
strength of the alloy in the as-built condition are common. The
working remediations are stress-relief heat treatment (annealing at
sub-recrystallisation temperatures), hot-isostatic-pressing (closes
internal porosity and homogenises the microstructure), and build-
strategy optimisation (scan patterns, support structures, build
orientation) to minimise the cumulative residual stress.

\engterm{Working AM alloys (production qualified, as of 2024)}.
Aerospace and medical AM has converged on a small set of alloys:
Ti-6Al-4V (medical implants, aerospace brackets), Inconel 718 and
625 (hot-section parts, fuel injectors), 17-4 PH stainless (general
metal AM), AlSi10Mg (lightweight structural; the aluminium alloy
the L-PBF process tolerates), maraging steel (tooling, dies),
cobalt-chrome (medical implants, dental). Each alloy has been
process-qualified for L-PBF or EBM at specific machine-and-parameter
combinations; substitution to a different machine or to a different
alloy with the same nominal chemistry requires re-qualification.

\engterm{Production examples}. GE Aviation's LEAP fuel-nozzle tip,
introduced in 2015, consolidated a 20-part assembly into a single
cobalt-chrome L-PBF part; the process now produces over $30{,}000$
parts per year and is the production proof point for AM in
flight-critical aerospace service. Stryker's titanium acetabular cups
(L-PBF) and Boeing's titanium environmental-control-system ducting
(EBM) are working examples of medical and aerospace AM at production
scale.

\section{Quality control: NDT methods}\pathtag{standard}

A manufactured part is qualified for service by the demonstration
that its defects are below the acceptable size and density limits.
Non-destructive testing (NDT) finds defects in a part without
destroying it. Six working methods cover the bulk of engineering
practice.

\engterm{Visual inspection (VT)}. The eye, with or without
magnification, sees surface-breaking defects of size approximately
$0.5$ -- $1\,\text{mm}$ and above on well-prepared surfaces. The
working limit for a fastener inspection or for a coating-defect
inspection; the first line of defence in a fabrication shop.

\engterm{Liquid penetrant testing (PT)}. A dye is applied to a
cleaned surface; the dye penetrates surface-breaking cracks by
capillary action; excess dye is removed; a developer draws the
remaining dye out of the cracks and onto the surface. Detection
limit $\sim 30\,\si{\micro\meter}$ wide cracks. Cheap, fast,
limited to surface-breaking defects, applies to all materials. The
canonical first NDT pass after machining or surface treatment.

\engterm{Magnetic particle testing (MT)}. A magnetic field is applied
to a ferromagnetic part; the field flux lines distort around
surface-breaking and near-surface (within $\sim 1\,\text{mm}$) cracks;
fine magnetic-particle powder applied to the surface accumulates at
the flux distortions and visualises the cracks. Detection limit $\sim
10\,\si{\micro\meter}$ wide cracks. Limited to ferromagnetic materials
(carbon and low-alloy steels). The canonical pass on welds in
structural steel.

\engterm{Eddy current testing (ET)}. An AC-driven coil produces an
oscillating magnetic field in the part; defects in the conducting
metal change the induced eddy currents; the change is detected as a
phase-and-amplitude shift in the coil's impedance. Detection of
surface and near-surface defects to depths of a few millimetres at
working frequencies of $\sim 100\,\text{kHz}$ -- $1\,\text{MHz}$;
deeper for lower frequencies. The canonical pass on aircraft skin
fasteners (the post-Aloha aircraft inspection regime relies heavily
on eddy current at fastener holes), heat-exchanger tubes, and
turbine-blade-cooling-passage walls.

\engterm{Ultrasonic testing (UT)}. A piezoelectric transducer sends
high-frequency ($1$ -- $20\,\text{MHz}$) sound pulses into the part;
reflections from internal defects and from the back wall are
received; the time-of-flight gives the defect's depth and the
amplitude gives an indication of its size. Detection of internal
defects to depths of metres at $1\,\text{MHz}$, fraction of a
millimetre at $20\,\text{MHz}$. The canonical pass on thick plate,
weld root-pass inspection, in-service condition monitoring of
pressure vessels.

\engterm{Radiographic testing (RT)}. X-ray or gamma-ray radiation
passes through the part; defects with different attenuation produce
contrast on a film or digital detector. The working method for
casting and welding inspection. Detection of volumetric defects
(porosity, inclusions, lack-of-fusion) to $\sim 1$ -- $2\,\%$ of
section thickness. Sensitive to thickness, less sensitive to planar
defects oriented parallel to the beam.

\engterm{Computed tomography (CT) and other modern methods}. Industrial
X-ray CT produces a three-dimensional image of the part's internal
structure at voxel sizes of $10$ -- $100\,\si{\micro\meter}$. Working
for AM-part qualification (CT reveals internal porosity that no other
NDT method finds with comparable fidelity), for composite-laminate
inspection, and for forensic examination of failed parts. Working
limitations: scan times of hours for high-resolution work, part-size
limits (typical industrial CTs handle parts up to $\sim 500\,\text{mm}$),
and metal-density limits (CT through steel above $\sim 50\,\text{mm}$
requires linear-accelerator X-ray sources).

\engterm{Acoustic emission, thermography, optical methods}. Acoustic-
emission monitoring detects the elastic waves released by growing
cracks (used for in-service monitoring of pressure vessels under
proof-test loading); active thermography excites a part thermally
and detects the temperature differences caused by internal flaws
(used on composite laminates and on aircraft skin); phased-array
ultrasound and laser-shearography are working in-service tools for
aerospace composites and for high-value forgings.

\engterm{Acceptance criteria}. NDT results are interpreted against
acceptance criteria drawn from the part's service requirements and
the relevant code (ASME B\&PV, AWS D1.1, ASTM E1255, MIL-STD-1530,
the specific aerospace SPC: Boeing 737 has its own NDT criteria,
Airbus the same). The criterion is the threshold above which the
part is rejected; the criterion is set by the residual-strength
calculation that the part must satisfy with the largest tolerable
defect.

\section{Failure: as-manufactured vs as-designed: when the gap killed people}\pathtag{core}

The gap between as-designed and as-manufactured properties is the
canonical failure pattern of manufacturing-driven failures. The
chapter's failure section names three documented cases.

\engterm{The Hyatt Regency walkway connection (1981)}. The walkways
were originally designed with a single continuous threaded rod from
the atrium ceiling, passing through the box-beam ends of both the
fourth-floor and second-floor walkways. Each connection carried only
its own walkway's load. During fabrication, the design was modified
to two separate rods (one ceiling-to-fourth-floor, one fourth-floor-
to-second-floor) for ease of installation. The modification doubled
the load on the fourth-floor walkway connection. The structural
engineer of record stamped the modified shop drawings without
recalculating the connection's capacity. The connection failed under
the design load of a tea-dance crowd; $114$ people died
\cite{acc:nbs-hyatt-1982}.

The canonical lesson: a fabrication-time modification can change a
connection's load by a factor of two; the change is the
manufacturing-design gap. The remediation is procedural: every
fabrication-time change must trigger a recalculation of the affected
connection; the shop drawing must reflect the change; the
engineering-of-record review must confirm the change. The gap killed
$114$ people because the procedural check failed.

\engterm{The Hubble primary mirror polishing error (1990)}. The HST
primary mirror's spherical-aberration error was traced to a
manufacturing error during the polishing of the mirror: the null
corrector (the reference optical system used to verify the mirror's
figure during polishing) was assembled with a $1.3\,\text{mm}$
spacing error \cite{acc:nasa-hubble-optical-systems-1990}. The
polished mirror conformed to the wrong reference; the error was
not detected because the same null corrector was used for the
quality-check measurements that should have detected the polishing
error. A second, independent measurement (an alternate null corrector,
a different metrology approach) was available but not used because
the project deemed the redundant check unnecessary on cost grounds.

The canonical lesson: the as-manufactured mirror conformed to the
specification; the specification it conformed to was the wrong
specification, and the metrology that should have caught the error
was self-referential. The remediation is the requirement of an
independent metrology check, traceable to a different reference,
before acceptance of a critical-tolerance manufactured part.

\engterm{Service failures of welded structures with HAZ defects}.
The pattern recurs across heavy-industry experience: a welded
structure (a pressure vessel, a bridge member, an offshore platform
joint, a pipeline weld) develops a service crack at a HAZ defect
that the original NDT inspection did not detect. The defect was
either below the inspection sensitivity, above the inspection limit
but missed by the inspector, or above the acceptance criterion but
accepted under a non-conformance disposition that did not adequately
account for the service environment.

The named cases are many and not all in the project's registry; the
working lesson is that the HAZ is the most likely region of service
failure in a welded structure, and the inspection regime must be
designed for the failure modes the HAZ supports (hydrogen cracking,
reheat cracking, stress-corrosion cracking, fatigue at HAZ stress-
concentrating discontinuities). Volume X returns to specific cases
in the systems-engineering context.

\engterm{The common structure of manufacturing-driven failures}.
Across the three patterns, the structure is the same: a design-intent
property that the manufacturing route was supposed to deliver; an
as-manufactured property that differed from the design intent; an
inspection or acceptance regime that failed to detect the difference;
a service load that the as-manufactured part could not carry. The
remediation is procedural at every step: a written specification, a
qualified process, a qualified inspector, a non-conformance approval
gate that recalculates the residual capacity.

\begin{principle}[The as-manufactured part is the engineered part]
A design that assumes the as-manufactured properties match the
nominal alloy properties is a design that has skipped a step. The
working properties are those of the specific microstructure produced
by the specific process; the working inspection is the inspection of
the as-manufactured part; the working acceptance is the acceptance
of the as-manufactured part against the as-designed criterion. Where
the procedure fails, the as-manufactured-versus-as-designed gap
becomes the failure mode.
\end{principle}

\begin{archetype}[Scaling and failure]
A manufacturing route exhibits its own scaling and failure laws. A
small coupon of an alloy made by a certain process is a poor
predictor of a full-scale part's properties: the residual stresses
scale with section thickness, the defect populations scale with
processed volume, the HAZ scales with weld pass count, the AM
anisotropy scales with build height. The chapter's working examples
all reduce to the same scaling: predict the full-scale property from
the small-scale coupon and from the process-route physics, and check
the prediction by inspection of the full-scale part. Where the
scaling check fails, the part fails in service.
\end{archetype}

\begin{estimation}
\textbf{An aerospace bracket is to be made by laser powder-bed fusion
of Ti-6Al-4V. The bracket has dimensions $200 \times 100 \times 50\,
\text{mm}$ and a build-direction height of $50\,\text{mm}$. Estimate
the residual stresses at the bracket's base after the build,
assuming a typical L-PBF thermal cycle.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
Each layer (thickness $t \sim 30\,\si{\micro\meter}$) cools from the
melt-pool peak ($\sim 1700\,\degC$) to the local build-chamber
temperature ($\sim 200\,\degC$, the typical L-PBF build-plate heated
to limit residual stress). The temperature drop per layer is $\Delta T
\sim 1500\,\si{\kelvin}$.

The thermal contraction strain of titanium over this drop is $\alpha
\Delta T \approx 9 \times 10^{-6} \times 1500 = 0.0135$. The elastic
estimate of the residual stress, if the cooled layer were fully
constrained by the substrate, is $E \alpha \Delta T \approx 113 \times
10^9 \times 0.0135 \approx 1.5\,\text{GPa}$. This is well above the
yield strength of Ti-6Al-4V ($\sim 900\,\text{MPa}$ in the
as-built condition); the cooling layer yields, and the maximum
residual stress is limited by yield strength.

Each layer adds a yield-strength-magnitude residual stress, but the
build does not accumulate these directly: each new layer reheats the
underlying material, partially relieving the previous layer's residual
stress. The net residual stress at the bracket's base after the full
$50\,\text{mm}$ build is the integrated result of $\sim 1700$ layers
of thermal cycling, accumulated over the build.

Working measurements on L-PBF Ti-6Al-4V parts of similar height give
peak as-built residual stresses of $\sim 500$ -- $800\,\text{MPa}$
at the part-substrate interface, of magnitude comparable to the
yield strength. Without stress relief, these residuals can cause
build-plate warpage, post-build distortion when supports are removed,
and reduced fatigue life in service.

The post-build stress-relief heat treatment at $\sim 800\,\degC$ for
$2\,\text{h}$ reduces the residual stresses to $\sim 100$ -- $200\,\text{MPa}$;
HIP at $920\,\degC$ and $100\,\text{MPa}$ for $2\,\text{h}$ closes
internal porosity, homogenises the microstructure, and reduces
residuals further. A fatigue-qualified L-PBF Ti-6Al-4V bracket in
aerospace service has had both treatments and a CT inspection.

The postmortem: the simple thermal-contraction estimate gives the
right order of magnitude. The actual stress field is more complex
(it depends on scan strategy, build-plate temperature, support-
structure design, post-build cooling rate), and the working response
is the full post-build heat-treatment-and-NDT cycle, not the design's
unaided trust in the printer.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Characterise the property variation in an additively
manufactured sample]
The reader obtains an additively manufactured part (3D-printed
polymer or metal coupon set; or characterises a small metal AM part
obtained from a service-bureau), measures its anisotropy in stiffness
and strength, and infers the manufacturing-driven property variation.
The deliverable is a fifteen- to twenty-five-page report.

\textbf{Track}. Physical measurement.
\textbf{Hazard class}: medium. Cutting, machining, and load-testing
of printed parts; safety glasses, hearing protection, and standard
shop precautions. For polymer FDM/FFF the project is low-hazard; for
metal AM coupons obtained externally, follow shop safety.

\textbf{Procedure}.

\begin{itemize}[leftmargin=*]
\item Print or obtain a set of $\sim 10$ -- $20$ coupon-shaped
  specimens, all of the same material, but oriented at different
  build directions: half in the XY plane (flat), half in the Z
  direction (vertical). For FDM/FFF: print PLA or PETG at $\sim 50$
  -- $80\,\%$ infill, $0.2\,\text{mm}$ layer height, with the long
  axis of the coupon either horizontal (XY) or vertical (Z).
\item Measure and record the coupon dimensions and the as-printed
  mass for each specimen.
\item Test each specimen in three-point bending or in tension (a
  hand-cranked load test with a force gauge and a deflection
  measurement; or a benchtop universal testing machine if available).
  Record peak load, deflection at peak, and visible failure mode.
\item Compute the apparent stiffness and the apparent strength of
  each coupon; tabulate by build direction.
\end{itemize}

\textbf{Analysis}.

\begin{itemize}[leftmargin=*]
\item Compute the mean and the standard deviation of stiffness and
  strength for each build direction.
\item Compute the anisotropy ratio (XY stiffness/Z stiffness; XY
  strength/Z strength).
\item Identify the failure mode (matrix-dominated: layer separation
  in Z-oriented coupons; fibre-dominated, i.e.\ filament-deposition-
  oriented in XY) by inspection of the fracture surface (a phone-
  camera macro shot suffices).
\item Compare the measured anisotropy ratio with the published values
  for the chosen material and process.
\end{itemize}

\textbf{Reflection ($1{,}500$ to $2{,}500$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
\item How large is the anisotropy ratio for the chosen material? How
  does it compare with the working ranges for L-PBF metal AM?
\item What aspects of the print parameters (layer height, infill,
  printing temperature, cooling rate) most influenced the measured
  anisotropy?
\item How would the design of a load-bearing part change to
  accommodate the anisotropy?
\item One feature of the measurement that the reader had not
  anticipated.
\end{itemize}

\textbf{Deliverable}. The coupon dimensions and masses, the
load-deflection data, the analysis, and the report.

\textbf{Time}. Four to eight hours of printing (most of which is
print time), four to six hours of measurement, four to six hours of
write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A steel forging is hot-rolled at $1100\,\degC$ and air-cooled to
ambient. Estimate the residual surface stress using the elastic
contraction approximation, assuming the surface reaches $300\,\degC$
when the interior is at $700\,\degC$, and the yield strength at the
contraction temperatures.
\end{exercise}

\begin{exercise}
A welded joint in 25\,mm A36 plate has a fully-yielded weld residual
stress at the toe. Compute the effective fatigue-life reduction by
adding the residual stress as a mean-stress shift to a Goodman
fatigue curve with $\sigma_{\text{endurance}} = 100\,\text{MPa}$ and
$\sigma_{\text{UTS}} = 400\,\text{MPa}$ for an applied alternating
stress of $50\,\text{MPa}$.
\end{exercise}

\begin{exercise}
Compute the case depth required to extend the fatigue life of a
$50\,\text{mm}$-diameter shaft from $10^5$ cycles to $10^7$ cycles by
shot peening. Assume the shot peening introduces a peak compressive
surface stress of $-500\,\text{MPa}$ at $0.2\,\text{mm}$ depth, and
the applied alternating stress is $200\,\text{MPa}$.
\end{exercise}

\begin{exercise}
A laser-powder-bed-fusion Inconel 718 part is built at $50\,\text{mm}$
height. Estimate the as-built residual stress at the substrate-
interface and the post-stress-relief residual stress, using the
thermal cycle parameters given in the chapter.
\end{exercise}

\begin{exercise}
A carburised gear tooth has a case depth of $1\,\text{mm}$ at
$60\,\text{HRC}$ over a core of $25\,\text{HRC}$. Compute the
Hertzian contact stress at the tooth flank under the working tooth
load and assess whether the case depth is adequate.
\end{exercise}

\begin{exercise}
A welded steel pressure vessel of $25\,\text{mm}$ wall thickness has
a HAZ with hardness $300\,\text{HV}$ over a parent metal of
$180\,\text{HV}$. Compute the implied yield-strength ratio between
HAZ and parent and assess the risk of hydrogen cracking.
\end{exercise}

\begin{exercise}
An aluminium 6061-T6 part is anodised with a $25\,\si{\micro\meter}$
hardcoat oxide. Estimate the through-thickness modulus change of the
combined system and the fatigue-life modification implied by the
anodised surface.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the relation $\sigma_{\text{res}} = E \alpha \Delta T / (1 -
\nu)$ for the residual surface stress of a thin cooled surface fully
constrained by an isothermal bulk substrate.
\end{exercise}

\begin{exercise}
Derive the Goodman correction for mean-stress effects on fatigue
life, and apply it to a welded joint with a yield-magnitude tensile
residual stress.
\end{exercise}

\begin{exercise}
Derive the rate of grain growth in austenite as a function of
temperature and time, using the parabolic growth law $d^2 = d_0^2 +
K t \exp(-Q/RT)$.
\end{exercise}

\begin{exercise}
Show that the build-direction yield-strength reduction in L-PBF parts
follows a Hall-Petch-type relation modified by the columnar grain
structure aspect ratio.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the time and energy required to anneal a $500\,\text{kg}$
forged-aluminium aerospace part for stress relief.
\end{exercise}

\begin{exercise}
Estimate the throughput of a laser-powder-bed-fusion machine for
production of Ti-6Al-4V brackets of $1\,\text{kg}$ each, using
typical L-PBF deposition rates and machine utilisation.
\end{exercise}

\begin{exercise}
Estimate the depth of an inspection-detection limit for surface-
breaking cracks in a steel weld using liquid-penetrant testing, and
compare with the limit for eddy-current testing.
\end{exercise}

\begin{exercise}
Estimate the X-ray dose required to inspect a $50\,\text{mm}$-thick
steel weld for porosity at a sensitivity of $1\,\%$ of section
thickness, using a typical industrial radiographic source.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a one-dimensional thermomechanical finite-difference
solver for the residual stress in a cooling steel plate. Compare
with the elastic-only approximation and identify the temperature
range where plastic relaxation dominates.
\end{exercise}

\begin{exercise}
Simulate the eddy-current response of a surface-breaking crack of
varying length and depth, using a one-dimensional analytical model
for a coil over a half-space. Identify the threshold crack size
detectable at a given inspection signal-to-noise ratio.
\end{exercise}

\begin{exercise}
Implement a layer-by-layer thermal model of an L-PBF build, computing
the residual stress accumulation as a function of build height and
scan strategy. Compare alternative scan strategies (raster, island,
checkerboard) on the resulting maximum residual stress.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify a heat-treatment schedule (preheat, interpass, post-weld)
for a $50\,\text{mm}$-thick high-strength steel pressure-vessel weld
to avoid hydrogen cracking. State the rationale for each step.
\end{exercise}

\begin{exercise}
Specify the surface treatment for a high-strength steel automotive
coil spring that must achieve $10^7$ fatigue cycles at an alternating
stress of $\pm 600\,\text{MPa}$. State the case-hardening or
shot-peening parameters.
\end{exercise}

\begin{exercise}
Specify the L-PBF process parameters (laser power, scan speed, hatch
spacing, build orientation, post-build heat treatment) for a
production Ti-6Al-4V aerospace bracket. State the rationale for
each choice.
\end{exercise}

\begin{exercise}
Specify the NDT inspection regime for a critical aerospace casting
in cast aluminium A356-T6. State the inspection methods, the
acceptance criteria, and the production-sampling frequency.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A welded steel pipeline cracks in service three months after
commissioning. The crack initiated at the weld toe and ran a few
millimetres before arrest. Diagnose the failure mode (hydrogen
cracking, hot cracking, fatigue, SCC) by the available evidence.
\end{exercise}

\begin{exercise}
An L-PBF aerospace bracket shows substantial distortion after support
structure removal. Diagnose the manufacturing-driven cause and the
intervention that would have prevented it.
\end{exercise}

\begin{exercise}
An automotive engine valve fails by surface cracking after $50{,}000\,
\text{km}$ of service. The crack initiated at the tip of the valve
and propagated to the seat. Diagnose the failure mode and the
manufacturing-driven aspect of the failure.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A forged aluminium 7050 aerospace structural part fails by stress-
corrosion cracking in service. The part's grain flow direction is
not aligned with the load direction. Identify the manufacturing-driven
contribution to the failure and the remediation.
\end{exercise}

\begin{exercise}
A composite-overwrapped pressure vessel ruptures during the proof
test. The vessel was made by filament winding with an autoclave cure.
Identify the manufacturing aspects most likely contributing and the
quality-control intervention that would prevent recurrence.
\end{exercise}

\begin{exercise}
A welded bridge member fails by fatigue at a stress concentration
near a stop-start in the longitudinal weld. The structure had been
qualified by NDT before placement. Identify the manufacturing-driven
failure mode and the inspection-regime modification that would
prevent recurrence.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A start-up's pitch claims an L-PBF process for high-strength
aerospace aluminium 7075 with as-built properties matching the wrought
T6 condition. Write a one-paragraph reply identifying the working
questions about solidification cracking, residual stress, and
qualification regime.
\end{exercise}

\begin{exercise}
A facility manager proposes to eliminate shot peening of automotive
coil springs to reduce production cost, on the argument that the
material's improved alloy chemistry achieves the same fatigue life
without the surface treatment. Write a one-paragraph evaluation
identifying the working test that would settle the proposal.
\end{exercise}

\begin{exercise}
A colleague proposes to qualify an AM-produced replacement for a
forged part by matching only the static tensile properties. Write a
one-paragraph reply identifying the additional property domains and
inspection-regime adjustments that the qualification must include.
\end{exercise}
