\chapter{Stress and strain}
\label{vol03:ch03}

\epigraph{Ut tensio, sic vis: as the extension, so the force.}{Robert
Hooke, \emph{Lectures de Potentia Restitutiva, or, Of Spring} (1678),
the decoded form of the 1676 anagram \emph{ceiiinosssttuv}
\cite{hist:hooke1678}.}

\chapmeta{Half-life: durable. Archetypes: balance, stability, failure
(material). Exercise target: 40. Project track: physical
instrumentation plus analysis (hybrid). Hazard class: low (cantilever
load testing with strain gauges). Reader path default: standard, with
sections explicitly tagged. Named cases: Hyatt Regency walkway
collapse (1981, cross-referenced).}

\noindent
Chapter 2 closed with the observation that the global statics of a
structure does not catch a joint failure; the joint's free-body
diagram does. This chapter takes the next step inward: the joint's
free-body diagram does not catch the material's failure, either; the
material's stress state does. We pass from the resultant force on a
cross-section to the distribution of force inside the section, and
from the deformation of a member to the deformation of an
infinitesimal element of material at any point inside the member. The
moves give the engineer the working vocabulary of mechanics of
materials: stress, strain, the elastic constants, the principal
stresses, the yield criterion, and the stress concentration that
turns an apparently safe member into a part that breaks.

\section{Internal forces become stresses}\pathtag{core}

A beam with an axial tensile force $N$ at a cross-section transmits
that force across the section by a distribution of force per unit
area. The local force per unit area is the \engterm{stress}. For a
straight prismatic member loaded uniaxially and at sufficient distance
from the load application points, the distribution is uniform across
the section, and the stress is simply
\[
\sigma = \frac{N}{A},
\]
the axial force divided by the cross-sectional area. The dimension is
$[\text{force}]/[\text{length}]^2$, with SI unit the pascal,
$1\,\si{\pascal} = 1\,\si{\newton\per\meter\squared}$.

The uniformity assumption requires distance from the load application
points. Saint-Venant's principle, established by Adh\'emar Jean
Claude Barr\'e de Saint-Venant in 1855 and now a standard result in
elasticity, states that the difference between two statically
equivalent loading distributions on a small region of a body's
boundary decays rapidly with distance from the loaded region. Far
from the load, the only thing the cross-section knows about the load
is its resultant. The standard engineering rule is that the
uniformity assumption holds at distances exceeding the largest cross-
section dimension from any concentrated load. Within that boundary
layer, the stress distribution is non-uniform, and the local geometry
matters; we return to that case in section 3.8.

The principle is worth stating as a rate, not only as a fact, because the
engineering value is in how fast the non-uniform part dies away. Write the
stress on a cross-section as the uniform resultant part plus a
self-equilibrating remainder that carries no net force or moment. Saint-
Venant's principle is the statement that the remainder decays roughly as
$\exp(-\pi x / w)$ with distance $x$ along the member and $w$ the
cross-section dimension, so at one width the self-equilibrating part has
fallen to about four percent of its value at the boundary and at two
widths to a fraction of a percent. Figure
\ref{fig:vol03ch03:saint-venant-decay} draws the recovery: three sections
at increasing distance from a concentrated load carry profiles that start
sharply peaked, smooth within one width, and become uniform beyond it. The
practical readings are two. First, a stress raiser, a bolt bearing, a
welded lug, a point support, disturbs only its own neighbourhood, so an
analyst may model the far field with the simple resultant stresses and
splice in a local solution near each load introduction. Second, the
principle sets the gauge-placement rule for the chapter's own project: a
strain gauge must sit at least one member-width from the clamp before its
reading matches slender-beam theory, because inside that width the clamp's
particular way of gripping still colours the strain. The same exponential
that reassures the far-field analyst warns the near-field experimenter.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-saint-venant-decay.tex}

For a beam under transverse load, the internal force at a cross-section
has three components (axial $N$, shear $V$, bending moment $M$ in two
dimensions; six components in three dimensions). Each component
produces its own stress distribution across the section:

\begin{itemize}[leftmargin=*]
  \item Axial $N$ produces uniform normal stress $\sigma = N/A$.
  \item Bending moment $M$ produces a linear normal stress
    distribution $\sigma = -M y / I$ for elastic bending of a slender
    beam, with $y$ measured from the neutral axis and $I$ the second
    moment of area about that axis (chapter 8 of this volume).
  \item Shear $V$ produces a shear stress distribution. For a
    rectangular section, the shear stress is parabolic with maximum
    at the neutral axis and zero at the top and bottom surfaces; the
    formula $\tau = VQ / (Ib)$ gives the local shear stress for any
    section (chapter 8).
\end{itemize}

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-axial-bending-distribution.tex}

The stress at a point under combined loading is the sum of the
contributions, by superposition. The summation is valid because
elasticity is linear in the regime this chapter treats; non-linear
material behaviour breaks superposition and is the subject of section
3.7.

A worked instance fixes the three distributions in mind. Take a
rectangular cantilever of width $b = 20\,\si{\milli\meter}$ and depth
$h = 40\,\si{\milli\meter}$, carrying a tip load $P = 2\,\text{kN}$ at
a free length $L = 0.5\,\si{\meter}$, with an additional axial pull
$N = 30\,\text{kN}$. At the clamped section the bending moment is
$M = PL = 1000\,\text{N}\cdot\text{m}$, the area is $A = bh = 800\,
\si{\milli\meter}^2 = 8\times 10^{-4}\,\si{\meter}^2$, and the second
moment of area is $I = bh^3/12 = (0.02)(0.04)^3/12 = 1.07\times
10^{-7}\,\si{\meter}^4$. The axial stress is $\sigma_N = N/A =
30{,}000/8\times 10^{-4} = 37.5\,\text{MPa}$, uniform across the
section. The bending stress at the top fibre ($y = +h/2 = 20\,\si{
\milli\meter}$) is $\sigma_M = M y / I = (1000)(0.02)/(1.07\times
10^{-7}) = 187\,\text{MPa}$ in tension on the bottom fibre and
$187\,\text{MPa}$ in compression on the top fibre. The combined
extreme-fibre stress on the tension side is $\sigma_N + \sigma_M =
37.5 + 187 = 224\,\text{MPa}$, on the compression side $37.5 - 187 =
-150\,\text{MPa}$. The axial pull shifts the whole linear bending
profile up by a constant, moving the zero-stress (neutral) axis off
the centroid. Figure \ref{fig:vol03ch03:combined-axial-bending-profile}
plots the three profiles through the depth: the flat axial contribution,
the linear bending contribution antisymmetric about the centroid, and
their sum, whose zero crossing sits at $y = -\sigma_N/(M/(I/y_{\max})) =
-37.5/187 = -0.20$ of the half-depth rather than at the centroid. This
superposition, three formulas added at a point, is
the daily arithmetic of mechanics of materials.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-combined-axial-bending-profile.tex}

\begin{estimation}
\textbf{Question.} Can a human stand on a single ordinary drinking
straw without crushing it? More precisely, what compressive stress
does an $80\,\si{\kilogram}$ person put on a polypropylene straw, and
how does it compare to the material's yield strength?

\textbf{Working.} A drinking straw has an outer diameter of about
$6\,\si{\milli\meter}$ and a wall thickness of about $0.15\,\si{\milli
\meter}$. The load-bearing cross-section is the annular wall area,
$A \approx \pi d t = \pi (6\times 10^{-3})(0.15\times 10^{-3}) \approx
2.8\times 10^{-6}\,\si{\meter}^2$, roughly $2.8\,\si{\milli\meter}^2$.
An $80\,\si{\kilogram}$ person weighs $W = mg \approx 785\,\text{N}$. If
all the weight went axially into one straw, the compressive stress
would be $\sigma = W/A \approx 785/2.8\times 10^{-6} \approx 280\,
\text{MPa}$. Polypropylene yields at roughly $30\,\text{MPa}$, so the
direct compressive stress alone is nearly ten times the yield stress;
the straw would crush. The estimate also undersells the real failure
mode: a thin-walled tube under axial load buckles (Chapter 8) at a
stress far below the crushing stress, so the straw collapses sideways
long before the wall material yields. The order-of-magnitude verdict,
the straw cannot carry a person, holds under every reasonable
assumption, and it
illustrates the chapter's first lesson: the same force produces a
manageable stress on a thick member and a fatal stress on a thin one,
because stress is force per unit area and the area is small.
\end{estimation}

A fourth internal action, torsion, produces its own shear stress
distribution. A torque $T$ on a circular shaft generates a
circumferential shear stress that grows linearly with radius,
$\tau(r) = Tr/J$, where $J = \pi d^4/32$ is the polar second moment of
area; the surface stress is $\tau_{\max} = 16T/(\pi d^3)$. Figure
\ref{fig:vol03ch03:torsion-distribution} shows the distribution.
Because the material near the axis carries almost no load, a hollow
shaft of the same outer diameter is far more efficient in torsion per
unit weight, which is why drive shafts and bicycle cranks are tubular.
Torsion recurs in the strain-energy methods of Chapter 5 and the
torsional vibrations of Chapter 6.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-torsion-distribution.tex}

The engineer's first move on a stressed body is therefore the same as
the statician's first move on a loaded structure: identify the
internal forces at each cross-section, then translate those into the
stress field at every point. The first translation is global to
cross-section (statics, chapter 2); the second is cross-section to
point (this chapter).

\section{Normal and shear stress}\pathtag{core}

At a point inside a loaded body, the state of stress is a tensor
quantity, not a vector. The reason is that the force per unit area at
a point depends on the orientation of the surface the area is taken
on. Imagine an infinitesimal cube of material at the point, aligned
with a chosen coordinate system. Each face of the cube has a force
per unit area associated with it; the components of that force per
unit area on each face are what we call the components of the stress
tensor.

We adopt the convention that the stress component $\sigma_{ij}$ is
the force per unit area in direction $j$ acting on the face whose
outward normal is in direction $i$. The diagonal components
($\sigma_{xx}$, $\sigma_{yy}$, $\sigma_{zz}$) are \engterm{normal
stresses}: they act along the outward normal of their face. The
off-diagonal components ($\sigma_{xy}$, $\sigma_{xz}$, $\sigma_{yz}$,
and so on) are \engterm{shear stresses}: they act parallel to their
face.

A second convention: tension is positive, compression negative for
normal stresses. For shear, the sign is set by the direction
convention on the coordinate axes; a positive $\sigma_{xy}$ on the
$+x$ face acts in the $+y$ direction.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-stress-element.tex}

The stress tensor in three dimensions has nine components arranged in
a $3 \times 3$ matrix:
\[
[\sigma] = \begin{pmatrix}
\sigma_{xx} & \sigma_{xy} & \sigma_{xz} \\
\sigma_{yx} & \sigma_{yy} & \sigma_{yz} \\
\sigma_{zx} & \sigma_{zy} & \sigma_{zz}
\end{pmatrix}.
\]
The tensor is symmetric: $\sigma_{ij} = \sigma_{ji}$. The symmetry
follows from moment equilibrium of the infinitesimal cube. If it
were not symmetric, the cube would carry a net moment and would not
be in equilibrium. The symmetry leaves six independent components in
three dimensions, three in two dimensions ($\sigma_{xx}$,
$\sigma_{yy}$, $\sigma_{xy}$).

The stress tensor is sometimes written as the \engterm{Cauchy stress
tensor}, after Augustin-Louis Cauchy, who established its tensor
nature in 1822. Cauchy's contribution was the recognition that the
force per unit area at a point on any surface through the point can
be obtained from the stress tensor by a linear operation: if $\hat{n}$
is the unit outward normal to the surface, the traction vector $\vec
{t}$ on that surface is
\[
\vec{t} = [\sigma] \hat{n},
\]
or in index notation, $t_i = \sigma_{ij} n_j$. This is \engterm{Cauchy's
formula}, and it is the bridge from the global concept of stress on a
particular section to the local concept of stress at a point. The
formula's compactness hides the work: stress is a tensor because the
question ``what is the force per unit area at this point'' is
ill-posed without specifying the surface orientation.

\begin{principle}[The stress at a point is a tensor]
At any point inside a loaded body, the local state of stress is a
symmetric second-rank tensor with six independent components in three
dimensions. The traction vector on any surface through the point is
obtained from the tensor by Cauchy's formula, $\vec{t} = [\sigma]
\hat{n}$. The tensor character of stress is what makes stress
transformation under change of coordinate axes a non-trivial
operation; the Mohr's-circle construction of section 3.5 is the
graphical version of the transformation rule.
\end{principle}

\begin{archetype}[Failure: the stress state as the gate to fracture]
The recurring archetype of this chapter is failure, in the specific
sense that a part fails not when its average load is high but when the
local stress state at some point reaches the material's capacity. The
stress tensor is the object that carries that local information. A
member can have a comfortable average stress and a fatal peak stress at
a fillet or a hole; the global free-body diagram of Chapter 2 cannot
see the difference, and the stress field of this chapter can. Every
failure case in this chapter, the de Havilland window corner, the Hyatt
walkway connection, the keyed shaft, reduces to the same move: find the
stress tensor at the worst point, reduce it to principal stresses,
evaluate the failure criterion, and compare to the allowable. The
archetype is that the worst point governs, and the worst point is
almost never where the average is.
\end{archetype}

\begin{mastery}
\textbf{The full tensor transformation and the three invariants.} In
three dimensions the stress components transform under a rotation $Q$
(an orthogonal matrix, $Q^{\mathsf{T}}Q = I$) by the rule
$[\sigma'] = Q\,[\sigma]\,Q^{\mathsf{T}}$, or in index notation
$\sigma'_{ij} = Q_{ip}Q_{jq}\sigma_{pq}$. This is the defining property
of a second-rank tensor and the reason stress is one. Three
combinations of the components are unchanged by any rotation, the
\engterm{stress invariants}:
\[
I_1 = \sigma_{kk}, \quad
I_2 = \tfrac{1}{2}(\sigma_{ii}\sigma_{jj} - \sigma_{ij}\sigma_{ij}),
\quad I_3 = \det[\sigma].
\]
The principal stresses are the roots of the characteristic cubic
$\lambda^3 - I_1\lambda^2 + I_2\lambda - I_3 = 0$, which is why the
principal stresses themselves are coordinate-independent. Splitting the
tensor into a hydrostatic part $p\,\delta_{ij}$ with $p = I_1/3$ and a
\engterm{deviatoric} part $s_{ij} = \sigma_{ij} - p\,\delta_{ij}$
separates the volume-changing stress from the shape-changing stress.
The von Mises criterion of section 3.6 is a statement about the second
invariant of the deviatoric tensor alone, $J_2 = \tfrac{1}{2}s_{ij}
s_{ij}$, with $\sigma_{\text{vM}} = \sqrt{3 J_2}$; this is why a purely
hydrostatic stress, however large, never causes von Mises yield. A
material under deep-ocean pressure or at the centre of a planet is
squeezed equally on all sides; the deviatoric part is zero, and a
ductile metal there does not yield no matter how high the pressure.
\end{mastery}

\section{Strain: small deformation theory}\pathtag{core}

A loaded body deforms. The relative motion of nearby material points
is the \engterm{strain}. We restrict attention to small deformations,
in which the displacement of any point is small compared to the
body's dimensions and the strains are small compared to unity (a
strain of $1$ would mean a body had stretched to twice its length;
in elastic engineering practice strains are typically $10^{-4}$ to
$10^{-3}$, well within the small-strain regime).

Consider a one-dimensional bar of length $L$ that is stretched to
length $L + \Delta L$ by axial loading. The \engterm{engineering
normal strain} is
\[
\epsilon = \frac{\Delta L}{L},
\]
dimensionless. Engineering strain is the standard small-strain
measure for elastic problems; it is sometimes called the
\engterm{nominal} strain to distinguish from the \engterm{true} (or
logarithmic) strain $\epsilon_{\text{true}} = \ln(L'/L)$ used in
large-deformation plasticity. For elastic strains the two coincide
to leading order.

In two and three dimensions, strain is defined locally as a tensor.
For a displacement field $\vec{u}(x, y, z)$, the small-strain tensor
is the symmetric part of the displacement gradient:
\[
\epsilon_{ij} = \frac{1}{2}\left(\frac{\partial u_i}{\partial x_j}
+ \frac{\partial u_j}{\partial x_i}\right).
\]
The diagonal components are \engterm{normal strains}: $\epsilon_{xx}
= \partial u_x / \partial x$ is the fractional change in length of an
element originally aligned with the $x$-axis. The off-diagonal
components are \engterm{shear strains}: $\epsilon_{xy} = \frac{1}{2}
(\partial u_x/\partial y + \partial u_y/\partial x)$ is the
infinitesimal rotation-symmetric part of the angle change between
two material lines originally along the $x$- and $y$-axes.

A common notational hazard: engineering shear strain $\gamma_{xy} = 2
\epsilon_{xy}$. The factor of two arises because engineering shear
strain measures the total angle change between two perpendicular
lines, while the tensor strain $\epsilon_{xy}$ measures half of
that. Hooke's law for shear stress is conventionally written $\tau =
G \gamma$, using engineering shear strain $\gamma$, not $2G
\epsilon_{xy}$. The reader will encounter both conventions and must
track which is in use.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-strain-deformation.tex}

The strain tensor is symmetric (by definition of the symmetric part
of the gradient) and has six independent components in three
dimensions, three in two dimensions. The antisymmetric part of the
displacement gradient is the local rigid-body rotation, which does
not produce strain; the strain tensor captures only the
deformation.

The split of the displacement gradient into its symmetric and
antisymmetric parts is worth carrying explicitly, because it is the
algebra behind the statement that rotation produces no strain. Any
square matrix decomposes uniquely into a symmetric and an antisymmetric
piece,
\[
\frac{\partial u_i}{\partial x_j} = \underbrace{\frac{1}{2}\left(
\frac{\partial u_i}{\partial x_j} + \frac{\partial u_j}{\partial x_i}
\right)}_{\epsilon_{ij}\ \text{(strain)}} + \underbrace{\frac{1}{2}
\left(\frac{\partial u_i}{\partial x_j} - \frac{\partial u_j}{\partial
x_i}\right)}_{\omega_{ij}\ \text{(rotation)}}.
\]
The antisymmetric part $\omega_{ij}$ has zeros on its diagonal and
equal-and-opposite off-diagonal entries, so in two dimensions it carries
the single number $\omega_{xy} = \tfrac{1}{2}(\partial u_x/\partial y -
\partial u_y/\partial x)$, the local rigid-body rotation angle of the
material element. A pure rotation moves every point but stretches no
line and shears no angle, so it must leave the strain tensor at zero,
and the decomposition delivers exactly that: the rotation lands entirely
in $\omega_{ij}$ and never reaches $\epsilon_{ij}$. The point is not
pedantic bookkeeping. A finite-element displacement solution, or a
digital-image-correlation strain map, computes the full gradient at each
point and must symmetrise it before reporting strain; reporting the raw
gradient would credit a rotating-but-undeformed region with a spurious
shear.

\subsection{Strain-displacement and the compatibility condition}

The six strain components in three dimensions are derived from only
three displacement components, so the strains are not independent: they
must be the symmetric gradient of \emph{some} single-valued displacement
field. A strain field invented by writing down six arbitrary functions
of position will, in general, fail to be the gradient of any continuous
displacement, and a body forced to carry it would tear or overlap. The
constraints that a strain field must satisfy to come from a continuous
displacement are the \engterm{compatibility conditions}. They are
obtained by eliminating the displacements from the strain-displacement
relations. In two dimensions the three strains $\epsilon_{xx}$,
$\epsilon_{yy}$, $\gamma_{xy}$ come from two displacements $u_x$, $u_y$,
leaving one compatibility equation,
\[
\frac{\partial^2 \epsilon_{xx}}{\partial y^2} + \frac{\partial^2
\epsilon_{yy}}{\partial x^2} = \frac{\partial^2 \gamma_{xy}}{\partial x
\,\partial y}.
\]
The derivation is direct. Differentiate $\epsilon_{xx} = \partial u_x/
\partial x$ twice in $y$ and $\epsilon_{yy} = \partial u_y/\partial y$
twice in $x$; both produce mixed third derivatives of the displacements,
\[
\frac{\partial^2 \epsilon_{xx}}{\partial y^2} = \frac{\partial^3 u_x}
{\partial x\,\partial y^2}, \quad \frac{\partial^2 \epsilon_{yy}}
{\partial x^2} = \frac{\partial^3 u_y}{\partial x^2\,\partial y}.
\]
The engineering shear $\gamma_{xy} = \partial u_x/\partial y + \partial
u_y/\partial x$ differentiated once in $x$ and once in $y$ gives the sum
of exactly those two third derivatives, so adding the first two and
matching the third closes the identity. The physical content is a
jigsaw-puzzle statement: the deformed pieces of the body must still fit
together without gaps or overlaps, and the compatibility equation is the
fitting condition written in derivatives. Compatibility is the second of
the two pillars of every elasticity problem. Equilibrium (section 3.1)
says the stresses must balance the loads; compatibility says the strains
must come from a continuous displacement; Hooke's law (next section)
links the two. A complete elasticity solution satisfies equilibrium,
compatibility, the constitutive law, and the boundary conditions at
once, and the Lam\'e thick-cylinder result and the Kirsch hole solution
quoted later in the chapter are exactly such solutions.

The two-dimensional case shown is the whole story only in two dimensions.
In three dimensions the single compatibility equation becomes six, coupling
all six strain components, of which three have the form of the plane
equation above and three mix the shear strains; only three of the six are
independent, a Bianchi-type identity reducing the count. Carrying all six
is rarely necessary in practice, because the more common route to a
solution works in stresses rather than strains. Substituting Hooke's law
into the compatibility equations and using equilibrium to eliminate the
body-force terms produces the \engterm{Beltrami-Michell} compatibility
equations, written entirely in the stress components, and a candidate
stress field is admissible only if it satisfies them along with
equilibrium and the boundary tractions. This is why an Airy stress
function, the device that automatically satisfies plane equilibrium by
construction, must still be biharmonic ($\nabla^4 \phi = 0$): the
biharmonic condition is compatibility in disguise. A stress field that
balances the loads but violates compatibility is not wrong arithmetic; it
is a field no continuous elastic body can actually hold, and the Beltrami-
Michell equations are the filter that rejects it. The engineer meeting a
guessed or interpolated field, from a coarse mesh, a hand-sketched flow
of force, or a separated-variable trial, checks compatibility before
trusting it, exactly as the worked compatibility test later in the chapter
does for a candidate strain field.

\section{Hooke's law and elastic constants}\pathtag{core}

For an isotropic linear elastic material, the stress tensor and the
strain tensor are proportional by a linear relation involving two
independent elastic constants. The generalised Hooke's law, in tensor
form, is
\[
\sigma_{ij} = \lambda \,\epsilon_{kk} \,\delta_{ij} + 2\mu
\,\epsilon_{ij},
\]
where $\lambda$ and $\mu$ are the Lam\'e constants, $\delta_{ij}$ is
the Kronecker delta, and the repeated index $k$ is summed (so
$\epsilon_{kk} = \epsilon_{xx} + \epsilon_{yy} + \epsilon_{zz}$ is
the trace of the strain tensor, also called the volumetric strain).

In engineering practice the Lam\'e constants are usually replaced by
two more directly measurable constants: the \engterm{Young's modulus}
$E$ and the \engterm{Poisson's ratio} $\nu$.

\begin{itemize}[leftmargin=*]
  \item Young's modulus $E$ is the proportionality constant in
    uniaxial loading: $\sigma = E \epsilon$ for a bar stretched
    axially. Units are pascals. Steel has $E \approx 200\,\text{GPa}$,
    aluminum alloys $\approx 70\,\text{GPa}$, structural wood (along
    grain) $\approx 10\,\text{GPa}$, concrete $\approx 30\,\text{GPa}$.
  \item Poisson's ratio $\nu$ is the negative of the ratio of
    transverse strain to axial strain in uniaxial loading: a bar
    stretched axially contracts laterally by $\nu \epsilon$. For most
    engineering metals $\nu \approx 0.3$; for cork, $\nu \approx 0$
    (cork shows almost no lateral contraction, which is why it
    inserts cleanly into bottles); for rubber, $\nu \approx 0.5$
    (rubber is nearly incompressible).
\end{itemize}

Poisson's ratio is the constant most often taken for granted and most
worth a second look, because its sign and its two limits carry real
material behaviour. Figure \ref{fig:vol03ch03:poisson-auxetic} sets three
materials side by side under the same axial stretch. A conventional
positive-$\nu$ material thins transversely as it stretches, conserving
something close to volume; at the upper limit $\nu \to 1/2$ the material
conserves volume exactly and is incompressible, which is why the bulk
modulus $K = E/[3(1-2\nu)]$ diverges there and why rubber, soft tissue,
and a fully plastic metal all behave as though squeezing them changes
their shape but never their volume. At $\nu = 0$ the transverse dimension
does not respond to axial load at all, the behaviour of cork that lets it
enter a bottle neck without bulging against the glass. A negative Poisson
ratio, long thought impossible, is real: an \engterm{auxetic} material
widens when pulled, built from re-entrant cellular geometry that unfolds
under tension rather than stretching along its members, and its lower
bound $\nu > -1$ is the condition that keeps the shear modulus $G =
E/[2(1+\nu)]$ positive. The admissible window $-1 < \nu < 1/2$ is
therefore not an algebraic accident but the statement that both moduli
stay positive, and each end of the window is a material an engineer can
hold: the incompressible solid at one end, the fully auxetic lattice at
the other. The consequence for design is that $\nu$ governs the coupling
between directions. A constrained material near $\nu = 1/2$ develops
enormous transverse pressure when compressed in a rigid cavity, the
mechanism that makes a rubber bushing act nearly rigid when it cannot
bulge, while a low-$\nu$ material in the same cavity transmits little side
pressure. The single number that looks like a footnote decides whether a
seal grips or slips.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-poisson-auxetic.tex}

The \engterm{shear modulus} $G$ relates shear stress to engineering
shear strain, $\tau = G \gamma$. The \engterm{bulk modulus} $K$
relates mean pressure to volumetric strain, $p = -K \epsilon_{kk}$.
For isotropic materials these are not independent:
\[
G = \frac{E}{2(1+\nu)}, \quad K = \frac{E}{3(1-2\nu)}.
\]
Two constants suffice for isotropic linear elasticity; any pair from
$(E, \nu, G, K, \lambda, \mu)$ may be used, and the others are
algebraically determined.

The two relations $G = E/[2(1+\nu)]$ and $K = E/[3(1-2\nu)]$ are not
quoted from a table; they fall out of the constitutive law applied to
two special stress states, and carrying the algebra once fixes both the
results and the reason the constants are linked. For the shear modulus,
load the material in pure shear $\tau$ on the $x$- and $y$-faces. The
in-plane principal stresses of pure shear are $\sigma_1 = +\tau$ and
$\sigma_2 = -\tau$ (the Mohr circle of pure shear is centred at the
origin with radius $\tau$). The principal strain along axis $1$ follows
from the plane Hooke's law,
\[
\epsilon_1 = \frac{1}{E}(\sigma_1 - \nu\sigma_2) = \frac{1}{E}(\tau +
\nu\tau) = \frac{\tau(1+\nu)}{E},
\]
and $\epsilon_2 = -\epsilon_1$ by symmetry. The engineering shear strain
on the original $45^\circ$ faces equals the difference of the principal
strains, $\gamma = \epsilon_1 - \epsilon_2 = 2\epsilon_1 = 2\tau(1+\nu)/
E$. Dividing the applied shear by this strain returns the shear modulus
directly,
\[
G = \frac{\tau}{\gamma} = \frac{\tau}{2\tau(1+\nu)/E} = \frac{E}{2(1+
\nu)},
\]
so $G$ is not a third free constant: it is fixed the moment $E$ and
$\nu$ are. For the bulk modulus, load the material hydrostatically,
$\sigma_{xx} = \sigma_{yy} = \sigma_{zz} = -p$. Each normal strain is the
direct term less the Poisson contraction from the other two,
\[
\epsilon = \frac{1}{E}\big[-p - \nu(-p - p)\big] = \frac{-p(1 - 2\nu)}
{E},
\]
the same for all three directions by symmetry, so the volumetric strain
is $\epsilon_{kk} = 3\epsilon = -3p(1-2\nu)/E$. The bulk modulus is the
pressure per unit volumetric compression,
\[
K = \frac{-p}{\epsilon_{kk}} = \frac{E}{3(1 - 2\nu)}.
\]
The two derivations also pin the physical range of Poisson's ratio. The
shear relation needs $\nu > -1$ to keep $G$ positive; the bulk relation
needs $\nu < 1/2$ to keep $K$ positive and finite. The admissible window
is $-1 < \nu < 1/2$, and the two ends are physical limits, not algebraic
curiosities: $\nu \to 1/2$ is the incompressible limit where $K \to
\infty$ and the material refuses to change volume (rubber, soft tissue,
metals in fully plastic flow), while a hypothetical $\nu < 0$ is an
auxetic material that thickens when stretched, realised in re-entrant
foams and certain engineered lattices.

\subsection{Strain energy density}

Stretching a material stores energy, and the stored energy is the area
under the stress-strain curve. For a unit volume loaded uniaxially from
zero to a stress $\sigma$ along the linear elastic line $\sigma = E
\epsilon$, the work done per unit volume is
\[
u = \int_0^{\epsilon} \sigma\,d\epsilon' = \int_0^{\epsilon} E
\epsilon'\,d\epsilon' = \frac{1}{2}E\epsilon^2 = \frac{1}{2}\sigma
\epsilon = \frac{\sigma^2}{2E},
\]
the triangle of figure \ref{fig:vol03ch03:strain-energy-triangle}. The
three forms are interchangeable through Hooke's law, and the factor of a
half is the hallmark of energy stored along a straight load line: the
average force over the stroke is half the peak force. For a general
stress state the strain energy density generalises to the sum over all
components,
\[
u = \frac{1}{2}\sigma_{ij}\epsilon_{ij} = \frac{1}{2}\left(\sigma_{xx}
\epsilon_{xx} + \sigma_{yy}\epsilon_{yy} + \sigma_{zz}\epsilon_{zz} +
\tau_{xy}\gamma_{xy} + \tau_{yz}\gamma_{yz} + \tau_{zx}\gamma_{zx}
\right),
\]
each normal stress paired with its normal strain and each shear stress
with its engineering shear strain. The energy splits cleanly into a
volume-changing part and a shape-changing part, the same split the von
Mises criterion exploits: substituting the hydrostatic and deviatoric
decomposition of section 3.2 gives
\[
u = \underbrace{\frac{\sigma_m^2}{2K}}_{\text{volumetric}} +
\underbrace{\frac{1}{4G}\,s_{ij}s_{ij}}_{\text{distortional}},
\]
with $\sigma_m$ the mean stress and $s_{ij}$ the deviatoric stress. The
distortional term is what von Mises sets a ceiling on; the volumetric
term, driven by the mean stress alone, never causes ductile yield, which
is why deep-ocean hydrostatic pressure stores energy in a steel block
without yielding it.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-strain-energy-triangle.tex}

The energy view is the engineer's quickest route to two daily quantities.
The first is the \engterm{modulus of resilience}, the recoverable energy
a material stores up to first yield, $u_R = \sigma_y^2/(2E)$, the green
triangle of the figure. A spring is a device built to make $u_R$ large
per unit volume, which is why spring steels combine a high yield stress
with an ordinary modulus, and why a high modulus alone, far from helping,
shrinks the energy a given stress can store. The second is the
\engterm{toughness}, the total area under the curve to fracture, the
energy absorbed before the material breaks; resilience lives in the
elastic triangle, toughness in the whole area, and a material can be
strong in one and poor in the other.

For an isotropic linear elastic material in three dimensions, Hooke's
law in the engineering constants is
\[
\epsilon_{xx} = \frac{1}{E}\left[\sigma_{xx} - \nu(\sigma_{yy} +
\sigma_{zz})\right],
\]
with analogous equations for $\epsilon_{yy}$ and $\epsilon_{zz}$,
and the shear relations $\gamma_{ij} = \tau_{ij} / G$ for $i \neq j$.
The form makes evident that a uniaxial stress in one direction
produces strain in all three coordinate directions: axial extension,
and transverse contraction by Poisson's ratio.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-tensile-test-fit.tex}

The elastic constants are not abstractions; they are measured numbers,
and the measurement is the tensile test. A standard coupon of known
cross-section is gripped at both ends and pulled at a controlled rate
while a load cell records force and an extensometer records elongation.
Force divided by original area gives engineering stress; elongation
divided by gauge length gives engineering strain. The slope of the
straight-line portion is Young's modulus. Figure
\ref{fig:vol03ch03:tensile-test-fit} shows a 6061-T6 aluminium coupon
whose elastic region returns $E \approx 69\,\text{GPa}$ from a
least-squares fit (\texttt{code/stress\_strain\_fit.py} on
\texttt{data/aluminum\_6061\_tensile.csv}). The same test, carried past
yield, gives the yield stress by the $0.2\,\%$ offset construction: draw
a line of slope $E$ shifted to a strain of $0.002$ and read off where it
crosses the measured curve. For this coupon the offset yield is about
$276\,\text{MPa}$, the handbook value for the alloy. Reference values
for several materials are tabulated in
\texttt{data/material\_properties.csv} and used in the exercises.

Hooke's 1678 statement of the law was scalar (extension proportional
to force in a spring); Cauchy's 1822 generalisation made it tensorial.
The conceptual unity is intact: stress is a linear function of strain
for small deformations of elastic materials. The chapter's discipline
is the precise statement and use of that linear function.

\section{Stress transformation, Mohr's circle}\pathtag{core}

The stress tensor at a point is the same physical quantity regardless
of the coordinate system used to express its components, but the
numerical values of the components change under rotation of the
coordinate axes. The transformation rule is the standard tensor
rotation rule. In two dimensions, with axes rotated by angle $\theta$
counterclockwise, the new components are
\begin{align*}
\sigma_{x'x'} &= \frac{\sigma_{xx} + \sigma_{yy}}{2} +
\frac{\sigma_{xx} - \sigma_{yy}}{2}\cos 2\theta + \sigma_{xy}\sin
2\theta, \\
\sigma_{y'y'} &= \frac{\sigma_{xx} + \sigma_{yy}}{2} -
\frac{\sigma_{xx} - \sigma_{yy}}{2}\cos 2\theta - \sigma_{xy}\sin
2\theta, \\
\sigma_{x'y'} &= -\frac{\sigma_{xx} - \sigma_{yy}}{2}\sin 2\theta +
\sigma_{xy}\cos 2\theta.
\end{align*}
The factor of $2$ inside the trigonometric arguments is the source of
much undergraduate confusion; the physical meaning is that the
stress tensor is invariant under rotation by $\pi$, not by $2\pi$, so
the transformation period is $\pi$ rather than $2\pi$.

There exists a particular orientation of axes at which the shear
stresses vanish and the normal stresses take their extreme values.
The corresponding axes are the \engterm{principal axes}; the normal
stresses on those axes are the \engterm{principal stresses}. In two
dimensions, the principal stresses are
\[
\sigma_{1,2} = \frac{\sigma_{xx} + \sigma_{yy}}{2} \pm \sqrt{\left(
\frac{\sigma_{xx} - \sigma_{yy}}{2}\right)^2 + \sigma_{xy}^2},
\]
with $\sigma_1 \geq \sigma_2$ by convention. The orientation of the
principal axes is given by
\[
\tan 2\theta_p = \frac{2 \sigma_{xy}}{\sigma_{xx} - \sigma_{yy}}.
\]
The two principal stresses and the angle $\theta_p$ together replace
the three components $(\sigma_{xx}, \sigma_{yy}, \sigma_{xy})$. The
principal-axis representation strips away coordinate-system
arbitrariness and presents the stress state in its intrinsic form.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-principal-rotation.tex}

\subsection{The principal stresses as an eigenproblem}

The principal-stress formula is not an isolated trick; it is the
two-dimensional case of the eigenvalue problem for the stress tensor,
and seeing it that way connects the hand formula to the three-dimensional
result and to the numerical routine a computer runs. A principal
direction is a direction $\hat{n}$ on which the traction is purely
normal, with no shear component: the traction vector points along
$\hat{n}$ itself. Cauchy's formula $\vec{t} = [\sigma]\hat{n}$ then reads
\[
[\sigma]\hat{n} = \sigma\,\hat{n},
\]
the defining equation of an eigenvector $\hat{n}$ with eigenvalue
$\sigma$. The principal stresses are the eigenvalues of the stress
tensor and the principal directions are its eigenvectors. Rewriting as
$([\sigma] - \sigma I)\hat{n} = 0$, a non-trivial $\hat{n}$ exists only
where the matrix is singular, $\det([\sigma] - \sigma I) = 0$, the
characteristic equation. In two dimensions the determinant is
\[
\det\begin{pmatrix} \sigma_{xx} - \sigma & \sigma_{xy} \\ \sigma_{xy} &
\sigma_{yy} - \sigma \end{pmatrix} = (\sigma_{xx} - \sigma)(\sigma_{yy}
- \sigma) - \sigma_{xy}^2 = 0,
\]
a quadratic in $\sigma$,
\[
\sigma^2 - (\sigma_{xx} + \sigma_{yy})\,\sigma + (\sigma_{xx}\sigma_{yy}
- \sigma_{xy}^2) = 0.
\]
The quadratic formula on this gives the principal-stress expression
exactly. The sum of roots is $\sigma_1 + \sigma_2 = \sigma_{xx} +
\sigma_{yy}$, the trace, which is why the Mohr-circle centre $(\sigma_{xx}
+ \sigma_{yy})/2$ is invariant under rotation; the product of roots is
$\sigma_1\sigma_2 = \sigma_{xx}\sigma_{yy} - \sigma_{xy}^2$, the
determinant, also invariant. Trace and determinant are the two
two-dimensional stress invariants, the numbers a rotation cannot change.
Figure \ref{fig:vol03ch03:principal-rotation} shows the geometric
content: the eigenproblem is the search for the element orientation on
which the off-diagonal shear entries rotate away to zero.

In three dimensions the same logic gives a cubic. The characteristic
equation $\det([\sigma] - \sigma I) = 0$ expands to
\[
\sigma^3 - I_1\,\sigma^2 + I_2\,\sigma - I_3 = 0,
\]
with the three invariants $I_1 = \sigma_{kk}$ (trace), $I_2$ the sum of
the principal two-by-two minors, and $I_3 = \det[\sigma]$, the same
$I_1, I_2, I_3$ named in the mastery box of section 3.2. The three real
roots are the three principal stresses $\sigma_1 \geq \sigma_2 \geq
\sigma_3$; the tensor's symmetry guarantees the roots are real and the
eigenvectors orthogonal, so the three principal directions form a mutually
perpendicular frame. A short worked instance fixes the cubic. Take the
stress tensor (in MPa)
\[
[\sigma] = \begin{pmatrix} 50 & 30 & 0 \\ 30 & -20 & 0 \\ 0 & 0 & 40
\end{pmatrix}.
\]
The third axis is already principal because its row and column carry no
shear, so $\sigma = 40\,\text{MPa}$ is one principal stress outright, and
the other two come from the $2\times 2$ block. That block has trace $50 +
(-20) = 30$ and determinant $(50)(-20) - 30^2 = -1000 - 900 = -1900$, so
its eigenvalues solve $\sigma^2 - 30\sigma - 1900 = 0$, giving $\sigma =
15 \pm \sqrt{15^2 + 1900} = 15 \pm \sqrt{2125} = 15 \pm 46.1$, that is
$61.1$ and $-31.1\,\text{MPa}$. Ordering the three, $\sigma_1 = 61.1$,
$\sigma_2 = 40$, $\sigma_3 = -31.1\,\text{MPa}$. The intermediate
principal stress here is the out-of-plane $40$, which Tresca discards and
von Mises keeps; the maximum shear is $(\sigma_1 - \sigma_3)/2 = (61.1 +
31.1)/2 = 46.1\,\text{MPa}$, on planes at $45^\circ$ to the first and
third principal axes. A general three-dimensional tensor with all nine
entries populated has no such free axis, and the cubic must be solved in
full, by a numerical eigenvalue routine in practice; the structure,
characteristic polynomial, three real roots, orthogonal eigenvectors, is
the same.

The maximum shear stress at the point is also computable directly:
\[
\tau_{\max} = \frac{\sigma_1 - \sigma_2}{2} = \sqrt{\left(
\frac{\sigma_{xx} - \sigma_{yy}}{2}\right)^2 + \sigma_{xy}^2},
\]
attained on planes oriented at $45^{\circ}$ to the principal axes.
The maximum shear is the basis of the Tresca yield criterion (section
3.6).

The \engterm{Mohr's circle} is a graphical construction that
encapsulates the transformation. Plot points on the $(\sigma, \tau)$
plane representing the stress on the $x$-face ($\sigma_{xx},
\sigma_{xy}$) and on the $y$-face ($\sigma_{yy}, -\sigma_{xy}$). The
two points are diametrically opposite on a circle whose centre is at
$((\sigma_{xx} + \sigma_{yy})/2, 0)$ and whose radius is $\tau_{
\max}$. Rotating the physical axes by $\theta$ corresponds to
rotating the diametrically-opposite points by $2\theta$ on the
circle, in the opposite sense (clockwise on the circle for
counterclockwise on the body). The intersections with the
horizontal axis give the principal stresses; the highest point on
the circle gives $\tau_{\max}$.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-mohr-circle.tex}

A worked construction is the fastest way to internalise the circle.
Take the stress state $\sigma_{xx} = 80\,\text{MPa}$, $\sigma_{yy} =
40\,\text{MPa}$, $\sigma_{xy} = 30\,\text{MPa}$. The centre is at
$C = (\sigma_{xx} + \sigma_{yy})/2 = 60\,\text{MPa}$ on the normal-stress
axis. The radius is $R = \sqrt{((80-40)/2)^2 + 30^2} = \sqrt{20^2 + 30^2}
= \sqrt{1300} = 36.06\,\text{MPa}$. The principal stresses are $\sigma_1
= 60 + 36.06 = 96.06\,\text{MPa}$ and $\sigma_2 = 60 - 36.06 = 23.94\,
\text{MPa}$. The principal direction follows from $\tan 2\theta_p =
2\sigma_{xy}/(\sigma_{xx} - \sigma_{yy}) = 60/40 = 1.5$, giving
$2\theta_p = 56.3^{\circ}$ and $\theta_p = 28.2^{\circ}$. The maximum
in-plane shear is $\tau_{\max} = R = 36.06\,\text{MPa}$, on planes at
$45^{\circ}$ to the principal axes. Figure
\ref{fig:vol03ch03:mohr-circle} plots the circle with the $x$-face
point $X = (80, 30)$ and the $y$-face point $Y = (40, -30)$ at opposite
ends of a diameter. The calculator
\texttt{code/mohrs\_circle.py} returns the same numbers and verifies
them on uniaxial and pure-shear states.

Mohr's circle is the engineer's mental machinery for stress
transformation. The numerical formulas above are the analytical
back-up; in practice a working stress analyst sketches the circle
in seconds and reads off the principal stresses and the maximum
shear without computation.

In three dimensions there are three principal stresses $\sigma_1
\geq \sigma_2 \geq \sigma_3$ obtained as the eigenvalues of the
stress tensor. The Mohr's circle generalises to three circles in
the same plane; the maximum shear is $(\sigma_1 - \sigma_3)/2$,
attained on planes oriented at $45^{\circ}$ to the first and third
principal axes.

\section{Yield and failure criteria}\pathtag{core}

A ductile material under stress remains elastic up to a threshold
stress called the \engterm{yield stress} $\sigma_y$. Above the
threshold, the material begins to deform plastically: irreversible
deformation accumulates with continued loading. The question that
yield criteria answer is, for a general state of stress (with
several non-zero components of the stress tensor), at what
combination of stresses does the material yield?

The yield stress is not itself a tensor: it is a scalar property of
the material, determined by a uniaxial tension test. The yield
criterion is a function of the stress tensor that, at the yield
threshold, equals $\sigma_y$. Two criteria are in standard
engineering use.

\subsection{Tresca: maximum shear stress}

The \engterm{Tresca criterion}, due to Henri Tresca's nineteenth-
century experiments on extrusion of metals, predicts yield when the
maximum shear stress reaches half the uniaxial yield stress:
\[
\tau_{\max} = \frac{\sigma_1 - \sigma_3}{2} = \frac{\sigma_y}{2},
\quad \text{or equivalently} \quad \sigma_1 - \sigma_3 = \sigma_y.
\]
Tresca is conservative (it predicts yield at a lower combined-stress
state than experiments show for most ductile metals) and simple. It
is the default yield criterion in many design codes, including the
ASME Boiler and Pressure Vessel Code. The physical content is direct:
a ductile metal yields by crystallographic slip on planes of maximum
shear, so a criterion built on the maximum shear stress speaks the
language of the mechanism. Tresca depends only on the largest and
smallest principal stresses and ignores the intermediate one
$\sigma_2$, which is both its simplicity and its inaccuracy: experiments
show the intermediate principal stress does affect yield, by up to the
fifteen percent that separates Tresca from von Mises. Where a hand
calculation must be defensible and slightly pessimistic, Tresca is the
faster and safer instrument; where weight is paid for in fuel or
payload, the conservatism is a cost.

\subsection{Von Mises: octahedral shear / distortion energy}

The \engterm{von Mises criterion}, proposed by Richard von Mises in
1913 and given a physical interpretation by Heinrich Hencky in 1924
as a distortion-energy criterion, predicts yield when the von Mises
equivalent stress reaches the uniaxial yield stress:
\[
\sigma_{\text{vM}} = \sqrt{\frac{1}{2}\left[(\sigma_1 - \sigma_2)^2
+ (\sigma_2 - \sigma_3)^2 + (\sigma_3 - \sigma_1)^2\right]} =
\sigma_y.
\]
For a general stress state with off-diagonal components,
\[
\sigma_{\text{vM}} = \sqrt{\frac{1}{2}\left[(\sigma_{xx} -
\sigma_{yy})^2 + (\sigma_{yy} - \sigma_{zz})^2 + (\sigma_{zz} -
\sigma_{xx})^2 + 6(\sigma_{xy}^2 + \sigma_{yz}^2 + \sigma_{zx}^2)
\right]}.
\]
Von Mises is less conservative than Tresca and matches experimental
data on ductile metals more closely. It is the default in most
structural-analysis software and most fatigue and finite-element
work. The two criteria agree exactly for uniaxial stress and for
pure shear scaled appropriately; they differ most in biaxial states.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-yield-loci.tex}

The geometry of the two criteria is most legible in the plane-stress
$(\sigma_1, \sigma_2)$ plane of figure \ref{fig:vol03ch03:yield-loci}.
Von Mises is an ellipse; Tresca is the hexagon inscribed in it. They
touch at the uniaxial points and at equibiaxial tension, and they
diverge most in pure shear, where Tresca predicts yield at $\sigma_y/2$
and von Mises at $\sigma_y/\sqrt{3} \approx 0.577\,\sigma_y$, a fifteen
percent difference. A worked comparison: a point in a pressure-vessel
wall has principal stresses $\sigma_1 = 150$, $\sigma_2 = 75$,
$\sigma_3 = 0\,\text{MPa}$. The Tresca equivalent is $\sigma_1 -
\sigma_3 = 150\,\text{MPa}$. The von Mises equivalent is
$\sqrt{\tfrac{1}{2}[(150-75)^2 + (75-0)^2 + (0-150)^2]} =
\sqrt{\tfrac{1}{2}[5625 + 5625 + 22500]} = \sqrt{16875} = 129.9\,
\text{MPa}$. Against a yield stress of $\sigma_y = 140\,\text{MPa}$,
Tresca declares the point yielded ($150 > 140$) while von Mises declares
it safe ($130 < 140$). The fifteen-megapascal gap between the two
verdicts is the conservatism of Tresca made arithmetic, and the
\texttt{code/yield\_criteria.py} routine evaluates both for any stress
tensor.

The choice between Tresca and von Mises depends on the application.
For pressure vessels and code-governed structural design, Tresca's
conservatism is a feature. For component design where weight is at
a premium (aerospace, transportation), von Mises's accuracy is
preferable. Most working engineers know both and choose by context.

\begin{principle}[Yield is a tensor question]
A scalar yield stress $\sigma_y$ is a material property obtained from
a uniaxial tension test. The yield criterion (Tresca, von Mises, or
another) is the function of the full stress tensor that, at the
yield threshold, equals $\sigma_y$. The principal stresses are the
intermediate computation. A design check against yield requires (1)
the stress tensor at every point of interest, (2) the principal
stresses at that point, (3) the chosen yield criterion evaluated at
those principal stresses, and (4) comparison to the material's
yield stress with the appropriate factor of safety.
\end{principle}

\section{Material non-linearity, plasticity preview}\pathtag{mastery}

The linear elasticity of the previous sections assumes a stress-strain
relation that is linear and reversible. Real materials depart from
that idealisation in several directions.

\engterm{Plasticity}: above the yield stress, ductile materials
deform irreversibly. The stress-strain curve in uniaxial tension
typically shows a linear elastic region up to $\sigma_y$, a brief
yield plateau (in some materials), and a strain-hardening region in
which the stress required to continue plastic deformation grows with
accumulated strain. The strain-hardening behaviour is captured by
hardening laws (Ramberg-Osgood, power-law, bilinear) fit to
experimental data. After the maximum (ultimate) stress, the
specimen necks and fractures; the strain at fracture is the
material's ductility.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-stress-strain-curve.tex}

Figure \ref{fig:vol03ch03:stress-strain-curve} contrasts the ductile
and brittle stress-strain curves on one axis. The ductile curve carries
the whole story of a metal's mechanical response: the elastic slope $E$,
the yield stress $\sigma_y$, the strain-hardening rise to the ultimate
stress $\sigma_u$, and the post-necking decline to fracture. The area
under the curve to fracture is the toughness, the energy per unit volume
absorbed before the material breaks. A brittle material has a high
strength but a tiny area under its curve; it is strong but not tough,
which is why a ceramic tile resists a static press yet shatters on a
dropped hammer. The distinction between strength (the height of the
curve) and toughness (the area under it) is one a designer must hold
separately; a material can be excellent in one and poor in the other.

The strain-hardening region needs a formula before it can be used in a
calculation, and the standard one is the \engterm{Ramberg-Osgood
relation}, which writes the total strain as an elastic part plus a plastic
part that grows as a power of the stress,
\[
\epsilon = \frac{\sigma}{E} + \alpha\,\frac{\sigma_0}{E}\left(\frac{\sigma}
{\sigma_0}\right)^{n},
\]
with $\sigma_0$ a reference stress near yield, $\alpha$ a fitting
constant of order one, and $n$ the \engterm{hardening exponent}. The
elastic term dominates at low stress and gives the straight initial slope
$E$; the plastic term, negligible until $\sigma$ approaches $\sigma_0$,
takes over above it and rounds the curve into the hardening region.
Figure \ref{fig:vol03ch03:ramberg-osgood} plots the law for a low and a
high exponent. A low $n$ (say $5$) gives a gradual, rounded knee, the
response of an annealed or low-carbon material that hardens steadily as it
deforms; a high $n$ (say $20$ and up) gives a sharp corner approaching the
\engterm{elastic-perfectly-plastic} idealisation, a flat yield plateau with
no hardening at all, which is the limit $n \to \infty$ and the model most
hand plasticity calculations assume. The exponent is read from a tension
test by plotting the plastic strain against stress on log-log axes, where
the power law becomes a straight line of slope $1/n$. The relation earns
its place because it is smooth and invertible, so a finite-element code
can evaluate stress from strain and strain from stress at every increment
without a discontinuity at yield, and because its two shape parameters
$\alpha$ and $n$ fit a real hardening curve with an accuracy the bilinear
model cannot reach. The same law underlies the strain-based fatigue
analysis of Volume X, where the cyclic stress-strain curve is a
Ramberg-Osgood fit to the stabilised hysteresis loops rather than to the
monotonic tension curve.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-ramberg-osgood.tex}

\engterm{Strain-rate sensitivity}: the yield stress and the
hardening law depend on the rate of straining. At low strain rates
(quasi-static, typical of structural service), strain-rate effects
are small. At high strain rates (impact, blast, shock), the yield
stress can be twice or more the static value, and the hardening law
changes character.

\engterm{Temperature dependence}: the elastic moduli, the yield
stress, and the hardening law all depend on temperature. Steel near
its melting point behaves as a viscous solid; aluminum at $400^{\circ}\
\mathrm{C}$ has lost a third of its room-temperature yield strength;
polymers transition from rubbery to glassy at temperatures specific
to the polymer.

\engterm{Anisotropy}: many engineering materials (wood along and
across grain, rolled metals, composites, single crystals) are not
isotropic. The elastic constants depend on direction; the yield
stress depends on direction; the failure mode depends on direction.
The isotropic constitutive law of section 3.4 is replaced by a
tensorial elastic-stiffness matrix with up to twenty-one independent
components. The count is worth carrying because it explains why the
isotropic case needed only two constants. Hooke's law in full generality
links the six stress components to the six strain components by a
$6\times 6$ stiffness matrix $C_{ij}$, so $\sigma_i = C_{ij}\epsilon_j$
in the engineering (Voigt) six-vector notation, and the matrix has
thirty-six entries. The symmetry of the strain-energy density, which
requires that the work done be path-independent, forces $C_{ij} =
C_{ji}$, cutting the count to the twenty-one independent constants of a
fully anisotropic (\engterm{triclinic}) material. Material symmetry cuts
it further at each step: an \engterm{orthotropic} material with three
mutually perpendicular symmetry planes, the class that covers wood, rolled
plate, and a unidirectional composite ply, keeps nine independent
constants; a \engterm{transversely isotropic} material, isotropic in one
plane, keeps five; and a fully isotropic material, symmetric in every
direction, collapses to the two of section 3.4. The practical weight of
the count is that an orthotropic ply needs nine numbers measured in nine
tests before its stress state can be judged, and the off-axis
transformation of the composite worked example is the reason a laminate is
never analysed on a single strength: the stiffness and the strength both
carry direction, and a load resolved onto the weak transverse axis governs
long before the strong fibre axis is troubled \cite[ch.~5]{text:chawla-composites}.

\engterm{Time-dependent behaviour}: at elevated temperatures, metals
creep (strain accumulates under sustained load); under cyclic
loading, they fatigue (cracks initiate and grow); in
chemically-aggressive environments, they corrode and stress-corrosion-
crack. Chapter 2 of Volume X treats fatigue and fracture mechanics
in depth; Chapter 3 of Volume X treats corrosion, wear, and
degradation.

The linear-elastic, isotropic, time-independent, isothermal,
small-strain model of sections 3.1-3.5 is the engineer's working
default. It is exact for many real materials in many real situations,
and a good first approximation for the rest. The corrections listed
above are added as needed; the discipline is knowing which correction
the situation requires.

\section{Stress concentration}\pathtag{core}

Saint-Venant's principle, cited in section 3.1, says that far from a
loaded region the only thing the body knows is the load's resultant.
The complement of the principle is that \emph{near} a discontinuity,
the local geometry matters very much. A hole in a plate, a fillet at
a step in shaft diameter, a keyway in a rotating shaft, a square
inside corner in a notched bar, a thread root on a fastener: each
of these produces a local stress field much larger than the nominal
stress on the gross section. The factor by which the local stress
exceeds the nominal is the \engterm{stress concentration factor}
$K_t$.

For a circular hole of radius $a$ in an infinite plate under remote
uniaxial tension $\sigma_0$, the classical Kirsch solution (1898)
gives the stress at the hole's edge as
\[
\sigma_{\max} = 3 \sigma_0,
\]
so $K_t = 3$. The factor of three is independent of the hole
diameter (so long as the hole is small compared to the plate's other
dimensions). The maximum occurs at the two points on the hole's edge
that lie on the diameter perpendicular to the load direction; the
edges parallel to the load actually experience a compressive stress
of magnitude $\sigma_0$.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-stress-concentration-hole.tex}

Figure \ref{fig:vol03ch03:stress-concentration-hole} plots the Kirsch
stress along the line perpendicular to the load, outward from the hole
edge. The amplification is sharp and local: it is $3\sigma_0$ at the
edge and within ten percent of the remote stress by three radii out.
The locality is Saint-Venant's principle running the other way, and it
is the reason a stress concentration is a point problem rather than a
section problem.

Other canonical concentration factors, from elasticity-theory
solutions and from experimental measurement
\cite[appendices]{text:crandall-mech-of-solids}:

\begin{itemize}[leftmargin=*]
  \item Elliptical hole with semi-axes $a$ (perpendicular to load)
    and $b$ (along load): $K_t = 1 + 2 a / b$. A circular hole has
    $a = b$ and $K_t = 3$; a long crack-like ellipse ($b \to 0$) has
    $K_t \to \infty$.
  \item Filleted shaft step (large to small diameter): $K_t$ ranges
    from about $1.5$ to $3$ depending on the ratio of fillet radius
    to small-shaft diameter; a sharp $90^{\circ}$ corner with zero
    radius has $K_t \to \infty$ in the elastic solution.
  \item Standard machine-screw thread root: $K_t \approx 2.5$ to
    $3.5$ in fatigue. The thread root is the canonical fatigue
    initiation site in a bolt.
  \item Square notch in a tension member, sharp internal corner:
    $K_t \approx 5$ or higher, depending on the depth ratio.
\end{itemize}

The elasticity factor $K_t$ is the linear elastic answer. Real
materials, especially ductile ones, redistribute stress locally by
small plastic deformation at the concentration site, and the
effective stress concentration in service is usually less than
$K_t$. For ductile static loading the redistribution is sufficient
that $K_t$ is often neglected and the gross-section stress used
instead. For brittle materials, for fatigue, and for dynamic
loading, $K_t$ is critical and the redistribution does not save
the part.

The distinction between the elastic factor and the factor a fatigue
calculation actually uses deserves its own name and rule. The elastic
$K_t$ is a property of geometry alone: it is what a linear-elastic
solution reports at the notch root and it makes no reference to the
material. The fatigue-effective factor $K_f$, the multiplier that
actually reduces the fatigue strength of a notched part, is usually
smaller, because a real material averages the stress over a small process
zone at the root rather than responding to the mathematical peak. The two
are bridged by the \engterm{notch sensitivity} $q$, defined so that
\[
K_f = 1 + q\,(K_t - 1), \qquad 0 \le q \le 1,
\]
with $q = 0$ a fully notch-insensitive material ($K_f = 1$, the notch does
nothing) and $q = 1$ a fully sensitive one ($K_f = K_t$, the notch bites
at full elastic strength). Notch sensitivity rises with notch root radius
and with material strength: a sharp notch in a hard, high-strength steel
approaches $q = 1$, while a generous fillet in a soft, ductile alloy can
fall well below it \cite[ch.~7]{text:dowling-mechanical-behavior}. The
practical error the rule guards against runs both ways. Applying the full
elastic $K_t$ to a blunt notch in a ductile alloy is over-conservative and
adds needless size and weight; ignoring the concentration entirely, the
gross-section temptation, misses the fatigue driver completely and ships a
part that cracks at the root. The correct discipline is to read $K_t$ from
the geometry, read $q$ from the root radius and the material, form $K_f$,
and apply $K_f$, not $K_t$, to the alternating stress. The shoulder-fillet
and bolted-joint worked cases later in the chapter use exactly this chain,
and Peterson's charts are the standard tabulation of $K_t$ from which it
starts \cite[introduction]{text:peterson-davie}.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-fillet-kt.tex}

The fillet is the designer's principal control over a concentration.
Figure \ref{fig:vol03ch03:fillet-kt} shows how $K_t$ at a stepped shaft
falls as the fillet radius grows: a sharp shoulder drives $K_t$ toward
the elastic singularity, while a fillet of radius $r > 0.1\,d$ brings it
under the usual design target of $2$. The cost of the fillet is a small
increase in part size at the transition; the cost of omitting it is a
fatigue crack at the shoulder. This is the trade the chapter keeps
returning to.

The engineering practice is to design fillet radii, fillets, and
transitions to keep $K_t$ low (typically below $2$) at every
geometric discontinuity. Sharp internal corners are the most
expensive things in mechanical design.

\section{Failure: where parts actually break and why}\pathtag{core}

A material loaded above the threshold its constitutive law allows
will fail. The chapter's failure section names the canonical sites
and mechanisms.

\subsection{Failure modes by material class}

\engterm{Ductile metals} (mild steel, aluminum alloys, copper) fail
by yielding followed by plastic deformation followed by ductile
fracture. The fracture surface is rough, often shows a cup-and-cone
geometry, and is preceded by visible necking. The yield criterion
governs the onset; the strain-hardening exponent governs the
ductility before fracture.

\engterm{Brittle materials} (cast iron, ceramics, glass, concrete in
tension, hardened steels) fail by sudden fracture with little or no
preceding plastic deformation. The fracture is initiated at a
microscopic flaw and propagates rapidly under tensile loading. The
governing model is fracture mechanics rather than yield: a stress
concentration that a ductile metal would redistribute, a brittle
material amplifies into crack propagation. Brittle materials are
strong in compression, weak in tension; their tensile strength is
typically $5$ to $20$ times less than their compressive strength.

\engterm{Composites} fail by a mixture of matrix yielding, matrix
cracking, fibre fracture, fibre-matrix debonding, and delamination,
in proportions that depend on the laminate stacking sequence and
the loading direction. The failure criterion is multi-component
(Tsai-Wu, Hashin, etc.) and the failure mode is often non-catastrophic
(damage accumulates gradually before the laminate's load capacity is
exceeded).

\engterm{Polymers} can be ductile (polycarbonate, polyethylene at
room temperature) or brittle (PMMA, polystyrene) depending on
temperature and strain rate; the same polymer can be ductile at low
strain rate and brittle at impact rates. Polymer failure is also
strongly time-dependent through creep at sustained stress and
through stress-relaxation at sustained strain.

\subsection{The stress concentration as failure initiator}

Real parts almost always fail at stress concentrations. The
canonical sites are bolt holes, fillet radii, thread roots, weld
toes, keyways, sharp internal corners, surface scratches, and
fabrication defects. Each of these is a place where the local
stress field is significantly larger than the nominal stress on
the gross section.

The mechanism is most acutely visible in two cases:

\engterm{Brittle fracture under static load}: a brittle material at
a sharp stress concentration sees a local stress that exceeds its
fracture strength even when the nominal stress is well below it.
The crack initiates at the concentration and propagates rapidly
across the section. Historical case: the early commercial jet
airliners of the mid-twentieth century experienced sudden in-flight
fuselage breakup after a small number of flight cycles. The accident
investigations identified fatigue cracks initiating at the square
corners of cabin window cutouts; the square corners had elastic
stress-concentration factors of approximately $K_t \approx 4$, and
the cyclic pressurisation produced fatigue at the concentration
even though the gross-section stresses were well below the
material's static strength. Subsequent airliner designs used
rounded (low-$K_t$) cabin windows; the lesson generalised across
all pressurised aerospace structures.

\engterm{Block shear at a bolted joint}: the modified upper
box-beam connection of the Hyatt Regency walkway, introduced in
chapter 1 of this volume, failed by block shear, the bolted-joint
failure mode of chapter 2. The mechanism is a stress concentration
at the bolt-to-plate interface combined with tension on the
plate's net section and shear along the planes parallel to the
load between the bolt and the plate edge. The NBS report
identified the doubled load on the modified connection and the
geometry of the bolt-to-flange spacing as the proximate causes;
the deeper cause is the stress concentration that any bolted
connection carries by construction, and the engineer's discipline
of designing the bolt pattern, edge distance, and plate thickness
to keep the concentration's amplified stress below the material's
capacity \cite{acc:nbs-hyatt-1982}.

\subsection{The diagnostic principle}

When a part has failed in service, the first diagnostic question is
where the fracture initiated. The initiation site is almost always
a stress concentration. The second question is whether the
concentration was anticipated in the design (and a finite, conservative
$K_t$ applied) or unanticipated (a defect, a fabrication scratch,
a service-induced notch, an unintended geometric feature). The
third question is whether the loading exceeded the design envelope.
The diagnostic discipline is the subject of Chapter 12 of Volume X,
forensic engineering; the chapter-3 lesson is that mechanics of
materials is the engineering of the answer to question one.

\begin{principle}[Parts break at stress concentrations]
Static or fatigue failure of a mechanical part initiates, in nearly
all cases, at a stress concentration: a hole, a fillet, a thread
root, a weld toe, a keyway, a surface scratch, or an internal defect.
Design against failure requires both (a) the stress field on the
gross section, computed from the loads and the section geometry, and
(b) the stress concentration factor at every geometric discontinuity,
applied as a multiplier on the nominal stress. The product of
nominal stress and $K_t$ is the local stress to be compared against
the material's allowable.
\end{principle}

\section{Worked examples}\pathtag{standard}

The single-action stress formulas of the preceding sections become
engineering only when several of them act at one point and the engineer
has to combine them, reduce the result to principal stresses, and check
it against a criterion. Three worked examples carry that arithmetic
through. Each begins from loads, builds the stress tensor at the worst
point, and ends with a numerical verdict. The reader who follows the
numbers rather than the words will own the method.

\subsection{Combined bending and torsion on a shaft surface}

A solid circular shaft of diameter $d = 30\,\si{\milli\meter}$ carries
a bending moment $M = 320\,\text{N}\cdot\text{m}$ and a torque $T =
210\,\text{N}\cdot\text{m}$ at the same cross-section. The worst point
is on the surface, where both the bending stress and the torsional
shear reach their maxima. We want the principal stresses and the von
Mises equivalent there, to compare against a yield stress $\sigma_y =
350\,\text{MPa}$.

The bending stress on the surface fibre is $\sigma_x = 32M/(\pi d^3)$.
With $d = 0.030\,\si{\meter}$, $\pi d^3 = \pi(0.030)^3 = 8.48\times
10^{-5}\,\si{\meter}^3$, so $\sigma_x = 32(320)/8.48\times 10^{-5} =
1.21\times 10^8\,\text{Pa} \approx 120\,\text{MPa}$. The torsional
shear on the same surface is $\tau = 16T/(\pi d^3) = 16(210)/8.48\times
10^{-5} = 3.96\times 10^7\,\text{Pa} \approx 40\,\text{MPa}$. The
surface element therefore carries a normal stress $\sigma_x = 120\,
\text{MPa}$ on its axial faces and a shear $\tau = 40\,\text{MPa}$ on
all four faces, with $\sigma_y = 0$ (the circumferential direction
carries no direct stress in pure bending plus torsion). The stress
state is the plane state $(\sigma_x, \sigma_y, \sigma_{xy}) = (120, 0,
40)\,\text{MPa}$.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-combined-loading-element.tex}

Figure \ref{fig:vol03ch03:combined-loading-element} shows the element
and its Mohr circle. The centre is $C = (\sigma_x + \sigma_y)/2 = 60\,
\text{MPa}$. The radius is $R = \sqrt{((\sigma_x - \sigma_y)/2)^2 +
\tau^2} = \sqrt{60^2 + 40^2} = \sqrt{5200} = 72.1\,\text{MPa}$. The
principal stresses are $\sigma_1 = 60 + 72.1 = 132\,\text{MPa}$ and
$\sigma_2 = 60 - 72.1 = -12\,\text{MPa}$, with $\sigma_3 = 0$ out of
plane. The principal direction follows from $\tan 2\theta_p =
2\tau/(\sigma_x - \sigma_y) = 80/120 = 0.667$, giving $\theta_p =
16.8^{\circ}$. The maximum in-plane shear is $\tau_{\max} = R = 72\,
\text{MPa}$.

The minor principal stress is compressive even though the only direct
stress applied was tensile bending; the torsion has rotated the
principal axes and pushed one principal stress below zero. That
compressive principal stress is what makes the combined state more
severe than bending alone. The von Mises equivalent is
\[
\sigma_{\text{vM}} = \sqrt{\sigma_x^2 + 3\tau^2} = \sqrt{120^2 +
3(40)^2} = \sqrt{14400 + 4800} = \sqrt{19200} = 139\,\text{MPa},
\]
using the reduced plane-stress form for a normal-plus-shear state, and
the same number falls out of the principal-stress form $\sqrt{\sigma_1^2
- \sigma_1\sigma_2 + \sigma_2^2} = \sqrt{132^2 - 132(-12) + 12^2} =
\sqrt{17424 + 1584 + 144} = \sqrt{19152} \approx 138\,\text{MPa}$ (the
small gap is rounding). The Tresca equivalent is the difference between the largest and
smallest principal stresses, $\sigma_1 - \sigma_2 = 132 - (-12) =
144\,\text{MPa}$, at or above the von Mises value as the maximum-shear
criterion always requires; the smallest principal stress here is the
in-plane $\sigma_2 = -12\,\text{MPa}$, not the out-of-plane
$\sigma_3 = 0$. Against $\sigma_y
= 350\,\text{MPa}$ the von Mises factor of safety is $350/139 = 2.5$,
comfortable. The bending stress alone, $120\,\text{MPa}$, would have
given a deceptive factor of safety of $2.9$; the torsion costs about
twenty percent of the margin, and an engineer who checked bending only
would have over-stated the safety. The lesson repeats the chapter's
refrain in a new place: the worst point is governed by the full stress
tensor, not by the largest single component.

\subsection{Thermal stress in a restrained bar}

A steel rail segment of length $L$ is clamped between two
effectively rigid abutments at $15\,{}^{\circ}\mathrm{C}$. The rail is
then heated by direct sun and friction to $75\,{}^{\circ}\mathrm{C}$, a
rise $\Delta T = 60\,\text{K}$. The expansion coefficient of steel is
$\alpha = 12\times 10^{-6}\,\si{\per\kelvin}$ and the modulus is $E =
200\,\text{GPa}$. We want the stress the restraint produces and a
comparison to yield.

If the bar were free, it would lengthen by the free thermal strain
$\epsilon_{\text{th}} = \alpha\,\Delta T = (12\times 10^{-6})(60) =
7.2\times 10^{-4}$. The rigid abutments forbid that lengthening, so
they impose a mechanical strain equal and opposite to the thermal
strain: $\epsilon_{\text{mech}} = -\epsilon_{\text{th}}$. The total
strain is zero (the bar cannot change length), and the stress is the
mechanical strain times the modulus,
\[
\sigma = E\,\epsilon_{\text{mech}} = -E\alpha\,\Delta T = -(200\times
10^9)(12\times 10^{-6})(60) = -1.44\times 10^8\,\text{Pa} = -144\,
\text{MPa},
\]
compressive. The stress does not depend on the bar length or its
cross-section; it depends only on the modulus, the expansion
coefficient, and the temperature change. A short rail and a long rail,
a thick one and a thin one, all reach the same $-144\,\text{MPa}$ under
the same $60\,\text{K}$ rise, so long as the restraint is rigid.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-thermal-stress.tex}

Figure \ref{fig:vol03ch03:thermal-stress} plots the restrained
compressive stress against temperature rise for steel and for 6061
aluminium. The steel line has slope $E\alpha = (200\times 10^3\,\text{
MPa})(12\times 10^{-6}\,\si{\per\kelvin}) = 2.4\,\text{MPa per K}$, so a
$60\,\text{K}$ rise gives $144\,\text{MPa}$, well below the $250\,\text{
MPa}$ yield marked on the figure. The aluminium line is shallower in
modulus but steeper in expansion; its product $E\alpha = (70\times
10^3)(23\times 10^{-6}) = 1.61\,\text{MPa per K}$ falls below steel's,
so an aluminium bar under the same restraint and the same temperature
change carries less thermal stress than a steel one, a result that
surprises engineers who expect the more-expansive metal to suffer more.
The figure also marks the steel yield-onset rise, $\Delta T = 250/2.4
\approx 104\,\text{K}$: a fully restrained steel member in a building
fire or a sun-trapped continuous rail reaches compressive yield near a
hundred-kelvin rise and then buckles sideways, which is the mechanism
behind the summer rail buckles railways guard against with expansion
joints and controlled rail-laying temperatures.

A partial restraint changes the arithmetic but not the principle. If
the abutments yield by a gap $g$ before they take load, the bar first
expands freely into the gap and only the remaining thermal strain is
suppressed: $\sigma = -E(\alpha\,\Delta T - g/L)$. The length $L$
re-enters through the gap term, so a longer bar is more forgiving of a
fixed gap. The design move is to size the gap or the expansion joint so
that the suppressed strain, and hence the stress, stays within the
allowable across the full service temperature range.

\subsection{Thin-walled vessel under pressure and torque}

A thin-walled cylindrical shaft-vessel of inner radius $r = 0.30\,\si{
\meter}$ and wall thickness $t = 6\,\si{\milli\meter}$ carries internal
pressure $p = 1.6\,\text{MPa}$ and, at the same time, transmits a
torque $T = 40\,\text{kN}\cdot\text{m}$ through its wall. We want the
stress state in the wall and the von Mises equivalent, to check against
$\sigma_y = 300\,\text{MPa}$.

The pressure produces a hoop stress and an axial stress. The hoop
stress is $\sigma_{\theta} = pr/t = (1.6\times 10^6)(0.30)/(0.006) =
8.0\times 10^7\,\text{Pa} = 80\,\text{MPa}$. The axial stress (closed
ends) is half of that, $\sigma_z = pr/(2t) = 40\,\text{MPa}$. The
torque produces a shear stress in the wall. For a thin closed tube the
shear flow is $q = T/(2A_m)$ with $A_m = \pi r^2$ the mean enclosed
area, so the wall shear is
\[
\tau = \frac{T}{2 A_m t} = \frac{T}{2\pi r^2 t} = \frac{40\times 10^3}
{2\pi (0.30)^2 (0.006)} = \frac{40\times 10^3}{3.39\times 10^{-3}} =
1.18\times 10^7\,\text{Pa} \approx 11.8\,\text{MPa}.
\]
The wall element carries $\sigma_{\theta} = 80\,\text{MPa}$,
$\sigma_z = 40\,\text{MPa}$, and a shear $\tau = 11.8\,\text{MPa}$
coupling the two. The in-plane principal stresses are
\[
\sigma_{1,2} = \frac{\sigma_{\theta} + \sigma_z}{2} \pm \sqrt{\left(
\frac{\sigma_{\theta} - \sigma_z}{2}\right)^2 + \tau^2} = 60 \pm
\sqrt{20^2 + 11.8^2} = 60 \pm 23.2,
\]
so $\sigma_1 = 83.2\,\text{MPa}$ and $\sigma_2 = 36.8\,\text{MPa}$,
with $\sigma_3 \approx 0$ (the thin-wall approximation drops the small
radial stress $-p$). The von Mises equivalent is
\[
\sigma_{\text{vM}} = \sqrt{\sigma_1^2 - \sigma_1\sigma_2 + \sigma_2^2}
= \sqrt{83.2^2 - 83.2(36.8) + 36.8^2} = \sqrt{6922 - 3062 + 1354} =
\sqrt{5214} = 72.2\,\text{MPa}.
\]
Against $\sigma_y = 300\,\text{MPa}$ the factor of safety is $300/72.2
= 4.2$. The torque here is modest relative to the pressure, so it
barely perturbs the hoop-dominated state; had the torque been an order
of magnitude larger, the shear term under the square root would have
dominated and rotated the principal axes far off the hoop direction.
The example shows the standard combined-loading discipline applied to a
pressure-vessel wall: assemble every stress component the loads
produce, reduce to principal stresses, evaluate the criterion once.

\subsection{Interference fit: contact pressure and hub hoop stress}

A steel gear hub is shrunk onto a solid steel shaft. The shaft radius is
$a = 25\,\si{\milli\meter}$; the hub outer radius is $b = 50\,\si{\milli
\meter}$; both are steel with $E = 200\,\text{GPa}$, $\nu = 0.3$. The hub
bore is machined undersize so that the diametral interference is
$\delta_d = 0.05\,\si{\milli\meter}$, a radial interference $\delta =
0.025\,\si{\milli\meter}$. We want the contact pressure at the interface,
the peak hoop stress in the hub, the friction torque the joint can carry,
and the temperature rise to assemble it by heat-shrinking.

The interference is squeezed out as elastic deformation of both members:
the hub bore stretches outward and the shaft compresses inward, and the
sum of the two radial movements equals the radial interference. For a
solid shaft inside a hub of the same material, the contact pressure that
results is
\[
p = \frac{E\,\delta}{a}\,\frac{b^2 - a^2}{2 b^2} = \frac{(200\times 10^9)
(0.025\times 10^{-3})}{0.025}\,\frac{0.050^2 - 0.025^2}{2(0.050)^2}.
\]
The leading group is $(200\times 10^9)(0.025\times 10^{-3})/0.025 =
2.0\times 10^8\,\text{Pa}$, and the geometric factor is $(0.0025 -
0.000625)/(2)(0.0025) = 0.001875/0.005 = 0.375$, so $p = 2.0\times 10^8
\times 0.375 = 7.5\times 10^7\,\text{Pa} = 75\,\text{MPa}$. Figure
\ref{fig:vol03ch03:shrink-fit-pressure} shows the geometry and the
stress directions at the interface.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-shrink-fit-pressure.tex}

The hub is a thick-walled cylinder under internal pressure $p$ with a
free outer surface, so its hoop stress is largest at the bore,
\[
\sigma_\theta^{\text{bore}} = p\,\frac{b^2 + a^2}{b^2 - a^2} = 75\,
\frac{0.0025 + 0.000625}{0.001875} = 75 \times 1.667 = 125\,\text{MPa},
\]
a tensile stress, well below the steel's yield of $350\,\text{MPa}$. The
solid shaft, squeezed uniformly on its outer surface by $p$, sits in a
uniform biaxial compression $\sigma_r = \sigma_\theta = -75\,\text{MPa}$
throughout, which no solid shaft has trouble carrying. The joint holds by
friction: the normal force on the contact cylinder is $p\,(2\pi a\,L)$
over a hub length $L = 40\,\si{\milli\meter}$, so with a friction
coefficient $\mu = 0.15$ the torque capacity is $T = \mu\,p\,(2\pi a L)\,a
= 0.15(75\times 10^6)(2\pi)(0.025)(0.040)(0.025) \approx 1.8\,\text{kN}
\cdot\text{m}$, and the cold press-on axial force is $\mu\,p\,(2\pi a L)
\approx 71\,\text{kN}$, a large press. Heat-shrinking avoids the press:
to grow the bore enough to clear the $0.05\,\si{\milli\meter}$ diametral
interference plus a small assembly clearance needs a temperature rise of
roughly $\Delta T \approx \delta_d/(\alpha\,d_{\text{bore}}) \approx
(0.07\times 10^{-3})/[(12\times 10^{-6})(0.050)] \approx 117\,\text{K}$,
within reach of an oven or an induction heater. The interference fit is
the chapter's thick-cylinder result and its thermal-strain result
working together: the interference sets the contact pressure through
elasticity, the pressure sets the hoop stress through the Lam\'e
solution, and the assembly itself is a controlled thermal strain.

\subsection{Shoulder fillet under rotating bending}

A stepped shaft carries rotating bending. The shaft steps from $D = 50\,
\si{\milli\meter}$ to $d = 40\,\si{\milli\meter}$ at a shoulder, and the
rotating bending moment at the shoulder is $M = 300\,\text{N}\cdot
\text{m}$, fully reversed once per revolution. The corrected endurance
limit of the steel is $S_e = 200\,\text{MPa}$. We want the fatigue factor
of safety, and we want to see what a sharp shoulder would do to it.

The nominal bending stress on the small-diameter section is $\sigma_
{\text{nom}} = 32M/(\pi d^3) = 32(300)/[\pi(0.040)^3] = 9600/(2.01\times
10^{-4}) = 4.77\times 10^7\,\text{Pa} \approx 47.7\,\text{MPa}$. Because
the bending is fully reversed, this is the alternating nominal stress and
the mean stress is zero; the fatigue check reduces to the alternating
stress against the endurance limit, with the fillet concentration
applied. A well-radiused fillet with $r/d = 0.1$ ($r = 4\,\si{\milli
\meter}$) at $D/d = 1.25$ has an elastic concentration of about $K_t =
1.65$ from the fillet charts of figure \ref{fig:vol03ch03:fillet-kt}. The
fatigue concentration is reduced from the elastic value by the notch
sensitivity $q$, here about $q = 0.9$ for a $4\,\si{\milli\meter}$ radius
in steel, so $K_f = 1 + q(K_t - 1) = 1 + 0.9(0.65) = 1.58$. The peak
alternating stress at the fillet is
\[
\sigma_a = K_f\,\sigma_{\text{nom}} = 1.58 \times 47.7 = 75.4\,\text{MPa},
\]
and the fatigue factor of safety is $S_e/\sigma_a = 200/75.4 = 2.65$,
comfortable for infinite life. A sharp shoulder tells a different story.
A near-zero fillet radius drives $K_t$ toward $2.4$, and with the same
notch sensitivity $K_f = 1 + 0.9(1.4) = 2.26$, so $\sigma_a = 2.26 \times
47.7 = 108\,\text{MPa}$ and the factor of safety falls to $200/108 =
1.85$. The shaft that was comfortable at $K_f = 1.58$ is marginal at
$K_f = 2.26$, and the only difference is a few millimetres of fillet
radius the drawing must call out and the machinist must cut. The
distinction between $K_t$ and $K_f$ matters here: applying the full
elastic $K_t$ would be over-conservative for a part that real material
notch-desensitises, while ignoring the concentration entirely, as the
gross-section temptation invites, would miss the fatigue driver
completely.

\subsection{Strain-gauge rosette reduction to principal stresses}

A point on a loaded steel surface carries an unknown plane stress
state. The loads do not give it directly, so it is measured with a
$0$--$45$--$90$ rectangular rosette of three gauges, sketched in figure
\ref{fig:vol03ch03:rosette-element}, reading $\varepsilon_a = 600\,
\mu\varepsilon$ at $0^\circ$, $\varepsilon_b = 400\,\mu\varepsilon$ at
$45^\circ$, and $\varepsilon_c = -100\,\mu\varepsilon$ at $90^\circ$. The
material is steel with $E = 200\,\text{GPa}$ and $\nu = 0.3$. We
reduce the three readings to the principal stresses and the von Mises
equivalent.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-rosette-element.tex}

The gauges at $0^\circ$ and $90^\circ$ read the normal strains directly,
$\varepsilon_{xx} = \varepsilon_a = 600\,\mu\varepsilon$ and $\varepsilon
_{yy} = \varepsilon_c = -100\,\mu\varepsilon$. The $45^\circ$ gauge fixes
the shear: the strain-transformation rule at $45^\circ$ gives
$\varepsilon_b = \tfrac{1}{2}(\varepsilon_{xx} + \varepsilon_{yy}) +
\tfrac{1}{2}\gamma_{xy}$, so
\[
\gamma_{xy} = 2\varepsilon_b - \varepsilon_{xx} - \varepsilon_{yy} =
2(400) - 600 - (-100) = 300\,\mu\varepsilon.
\]
The in-plane strain tensor is $(\varepsilon_{xx}, \varepsilon_{yy},
\gamma_{xy}) = (600, -100, 300)\,\mu\varepsilon$. The principal strains
follow from the same Mohr construction the stresses use, with
$\gamma/2$ on the shear axis:
\[
\varepsilon_{1,2} = \frac{\varepsilon_{xx} + \varepsilon_{yy}}{2} \pm
\sqrt{\left(\frac{\varepsilon_{xx} - \varepsilon_{yy}}{2}\right)^2 +
\left(\frac{\gamma_{xy}}{2}\right)^2} = 250 \pm \sqrt{350^2 + 150^2} =
250 \pm 381\,\mu\varepsilon,
\]
so $\varepsilon_1 = 631\,\mu\varepsilon$ and $\varepsilon_2 = -131\,
\mu\varepsilon$, with the principal direction at $\theta_p = \tfrac{1}{2}
\arctan[300/(600-(-100))] = \tfrac{1}{2}\arctan(0.429) = 11.6^\circ$ from
the $x$-axis. The free surface is a plane-stress state ($\sigma_3 = 0$
out of plane), so the principal stresses come from the plane-stress
Hooke's law in principal form,
\[
\sigma_1 = \frac{E}{1-\nu^2}(\varepsilon_1 + \nu\varepsilon_2), \quad
\sigma_2 = \frac{E}{1-\nu^2}(\varepsilon_2 + \nu\varepsilon_1).
\]
With $E/(1-\nu^2) = 200\times 10^9 / (1 - 0.09) = 2.20\times 10^{11}\,
\text{Pa}$, the bracketed strain group for the major principal stress is
$\varepsilon_1 + \nu\varepsilon_2 = (631 + 0.3(-131))\times 10^{-6} =
592\times 10^{-6}$, so $\sigma_1 = 2.20\times 10^{11}(592\times 10^{-6})
= 1.30\times 10^8\,\text{Pa} = 130\,\text{MPa}$, and the minor principal
stress is $\sigma_2 = 2.20\times 10^{11}(\varepsilon_2 + \nu\varepsilon_1)
= 2.20\times 10^{11}(-131 + 0.3(631))\times 10^{-6} = 2.20\times 10^{11}
(58\times 10^{-6}) = 13\,\text{MPa}$, both tensile. The von Mises equivalent is
$\sigma_{\text{vM}} = \sqrt{\sigma_1^2 - \sigma_1\sigma_2 + \sigma_2^2} =
\sqrt{130^2 - 130(13) + 13^2} = \sqrt{16{,}900 - 1690 + 169} = \sqrt{15
{,}379} = 124\,\text{MPa}$. The rosette has measured a stress state no
load calculation supplied, and the reduction has turned three voltage
readings into a pair of principal stresses and a yield margin. This is the
experimental face of the chapter: where geometry and loading are too
tangled for a formula, the surface strain is measured and Hooke's law
runs in reverse, from strain to stress.

\subsection{Inverting the generalized 3D Hooke's law}

The three worked examples above all read strain from stress or stress
from strain in special states (uniaxial, plane stress, equibiaxial). A
general triaxial state requires the full inversion, and the inversion is
worth carrying once with numbers so the algebra is concrete. A point
deep inside a loaded steel forging carries the stress state $\sigma_{xx}
= 180\,\text{MPa}$, $\sigma_{yy} = 90\,\text{MPa}$, $\sigma_{zz} =
-60\,\text{MPa}$, with no shear on these axes (they happen to be the
principal axes). The steel has $E = 200\,\text{GPa}$ and $\nu = 0.3$. We
want all three normal strains, the volumetric strain, and the mean
(hydrostatic) stress, and we want to confirm the two routes to the
volumetric strain agree.

The generalized Hooke's law in engineering constants gives each normal
strain as the direct term less the Poisson contraction from the other
two,
\[
\epsilon_{xx} = \frac{1}{E}\left[\sigma_{xx} - \nu(\sigma_{yy} +
\sigma_{zz})\right], \quad \text{and cyclically for } yy, zz.
\]
With $E = 200\times 10^3\,\text{MPa}$ and $\nu = 0.3$, the $x$-strain is
\[
\epsilon_{xx} = \frac{1}{200{,}000}\left[180 - 0.3(90 - 60)\right] =
\frac{180 - 9}{200{,}000} = \frac{171}{200{,}000} = 855\,\mu\epsilon,
\]
the $y$-strain is $\epsilon_{yy} = [90 - 0.3(180 - 60)]/200{,}000 =
(90 - 36)/200{,}000 = 270\,\mu\epsilon$, and the $z$-strain is
$\epsilon_{zz} = [-60 - 0.3(180 + 90)]/200{,}000 = (-60 - 81)/200{,}000 =
-705\,\mu\epsilon$. The forging stretches along $x$ and $y$ and shortens
along $z$, as the signs of the stresses and the Poisson coupling
together demand. The volumetric strain is the sum,
\[
\epsilon_{kk} = \epsilon_{xx} + \epsilon_{yy} + \epsilon_{zz} =
(855 + 270 - 705)\,\mu\epsilon = 420\,\mu\epsilon.
\]
The second route to the same number runs through the bulk modulus. The
mean stress is $\sigma_m = (\sigma_{xx} + \sigma_{yy} + \sigma_{zz})/3 =
(180 + 90 - 60)/3 = 70\,\text{MPa}$, and the bulk modulus is $K =
E/[3(1-2\nu)] = 200{,}000/[3(0.4)] = 166{,}700\,\text{MPa}$. The
volumetric strain from the bulk relation is $\epsilon_{kk} = \sigma_{kk}
/(3K)$, but $\sigma_{kk} = 3\sigma_m = 210\,\text{MPa}$, so $\epsilon_{kk}
= 210/166{,}700 = 1260\,\mu\epsilon$. The two routes disagree by a factor
of three, which is the warning the careful reader should heed: the bulk
relation is $p = -K\epsilon_{kk}$ with $p = -\sigma_m$, so
$\epsilon_{kk} = \sigma_m/K = 70/166{,}700 = 420\,\mu\epsilon$, matching
the direct sum once $\sigma_m$ rather than $\sigma_{kk}$ is divided by
$K$. The factor-of-three trap, summing the three stresses where the mean
was meant, is one of the more common arithmetic slips in three-
dimensional elasticity, and the two-route check catches it every time.
The example also separates the response cleanly: the mean stress
$70\,\text{MPa}$ drives the $420\,\mu\epsilon$ volume change through $K$,
while the deviatoric stresses, $\sigma_{xx} - \sigma_m = 110$, $\sigma_
{yy} - \sigma_m = 20$, $\sigma_{zz} - \sigma_m = -130\,\text{MPa}$, drive
the shape change through $G$, and the two parts of the response never
cross-talk in an isotropic material.

\subsection{Octahedral shear and the von Mises criterion derived}

Section 3.6 stated the von Mises equivalent stress as a formula. The
formula has a physical origin worth carrying in full, because it shows
why von Mises is a statement about distortion alone and why the
$\sqrt{3}$ that separates it from a naive shear criterion appears. The
\engterm{octahedral plane} at a point is the plane whose normal makes
equal angles with all three principal axes; its unit normal is $\hat{n}
= (1, 1, 1)/\sqrt{3}$. There are eight such planes (the faces of a
regular octahedron inscribed in the principal-axis frame), and they
share one shear stress, the octahedral shear stress $\tau_{\text{oct}}$.

The traction on the octahedral plane follows from Cauchy's formula
$\vec{t} = [\sigma]\hat{n}$ applied in the principal frame, where the
stress tensor is diagonal with entries $\sigma_1, \sigma_2, \sigma_3$.
The traction components are $t_i = \sigma_i n_i = \sigma_i/\sqrt{3}$, so
the traction vector is $\vec{t} = (\sigma_1, \sigma_2, \sigma_3)/\sqrt{3}$.
Its magnitude squared is $|\vec{t}|^2 = (\sigma_1^2 + \sigma_2^2 +
\sigma_3^2)/3$. The normal component of the traction is the projection on
$\hat{n}$,
\[
\sigma_{\text{oct}} = \vec{t}\cdot\hat{n} = \frac{\sigma_1 + \sigma_2 +
\sigma_3}{3} = \sigma_m,
\]
the mean stress, as expected for a plane equally inclined to all three
axes. The shear component is what is left over by Pythagoras,
\[
\tau_{\text{oct}}^2 = |\vec{t}|^2 - \sigma_{\text{oct}}^2 =
\frac{\sigma_1^2 + \sigma_2^2 + \sigma_3^2}{3} - \left(\frac{\sigma_1 +
\sigma_2 + \sigma_3}{3}\right)^2.
\]
Expanding the square and collecting terms turns the difference into a
sum of squared principal-stress differences,
\[
\tau_{\text{oct}} = \frac{1}{3}\sqrt{(\sigma_1 - \sigma_2)^2 +
(\sigma_2 - \sigma_3)^2 + (\sigma_3 - \sigma_1)^2}.
\]
The octahedral shear depends only on the differences of the principal
stresses, never on their common part, so a hydrostatic stress (all three
equal) gives $\tau_{\text{oct}} = 0$. The von Mises equivalent stress is
this octahedral shear scaled to read $\sigma_y$ in uniaxial tension. In
uniaxial tension $\sigma_1 = \sigma$, $\sigma_2 = \sigma_3 = 0$, so
$\tau_{\text{oct}} = \sqrt{2}\,\sigma/3$; demanding that the scaled
quantity equal $\sigma$ in that state fixes the scale factor at
$3/\sqrt{2}$, and
\[
\sigma_{\text{vM}} = \frac{3}{\sqrt{2}}\,\tau_{\text{oct}} =
\sqrt{\frac{1}{2}\left[(\sigma_1 - \sigma_2)^2 + (\sigma_2 - \sigma_3)^2
+ (\sigma_3 - \sigma_1)^2\right]},
\]
the formula of section 3.6, now derived rather than asserted. Figure
\ref{fig:vol03ch03:von-mises-octahedral} draws the same content in
principal-stress space: on the deviatoric plane perpendicular to the
hydrostatic axis, von Mises is a circle of radius proportional to
$\tau_{\text{oct}}$, and Tresca is the inscribed hexagon. The $\sqrt{3}$
that separates the two criteria in pure shear is the ratio of the
hexagon's flat-to-flat to corner-to-corner dimensions, made physical:
the same deviatoric magnitude reaches yield sooner in the shear
directions, where Tresca's flats cut in, than in the uniaxial
directions, where the two loci touch.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-von-mises-octahedral.tex}

\subsection{Thermal mismatch in a two-material bonded bar}

The restrained-bar example heated one material between rigid walls. A
bonded pair of materials needs no external wall: each restrains the
other, and a uniform temperature change alone builds stress, because the
two materials want to expand by different amounts and the bond forbids
them to slide. A copper bar and a steel bar of equal cross-sectional
area $A = 100\,\si{\milli\meter}^2$ are bonded end to end into a single
bar clamped between two rigid supports, then heated by $\Delta T =
80\,\text{K}$. Copper has $E_c = 120\,\text{GPa}$, $\alpha_c = 17\times
10^{-6}\,\si{\per\kelvin}$; steel has $E_s = 200\,\text{GPa}$, $\alpha_s
= 12\times 10^{-6}\,\si{\per\kelvin}$. We want the internal force the bar
carries and the stress in each segment.

The two segments are in series between rigid walls, so they share one
internal axial force $F$ (compressive, since both want to lengthen and
cannot), and the total length change is zero. Each segment's length
change is its free thermal expansion plus its mechanical (force-driven)
change. For segment lengths $L_c$ and $L_s$, the zero-total-length
condition is
\[
\underbrace{\alpha_c \Delta T\,L_c - \frac{F L_c}{E_c A}}_{\text{copper}}
+ \underbrace{\alpha_s \Delta T\,L_s - \frac{F L_s}{E_s A}}_{\text{steel}}
= 0.
\]
The compressive force is therefore
\[
F = \Delta T\,A\,\frac{\alpha_c L_c + \alpha_s L_s}{L_c/E_c + L_s/E_s}.
\]
For equal segment lengths $L_c = L_s = L$ the length cancels, and the
expression reduces to a weighted combination of the two materials,
\[
F = \Delta T\,A\,\frac{\alpha_c + \alpha_s}{1/E_c + 1/E_s} = \Delta T\,A
\,\frac{(\alpha_c + \alpha_s)E_c E_s}{E_c + E_s}.
\]
Putting in numbers, the numerator group is $(\alpha_c + \alpha_s)E_c E_s
= (29\times 10^{-6})(120\times 10^9)(200\times 10^9) = 6.96\times
10^{17}$, the denominator $E_c + E_s = 320\times 10^9$, so the modulus
combination is $2.175\times 10^6\,\text{Pa}\cdot\si{\per\kelvin}$. Then
$F = (80)(100\times 10^{-6})(2.175\times 10^6) = 1.74\times 10^4\,\text
{N} = 17.4\,\text{kN}$, compressive. The stress is the same in both
segments because the area is the same, $\sigma = F/A = 17.4\times 10^3 /
100\times 10^{-6} = 174\,\text{MPa}$, compressive, in each. The bonded
bar reaches a substantial stress from temperature alone, which is the
mechanism behind thermal-mismatch failure in soldered joints, in
glass-to-metal seals, and in the copper-on-laminate of a printed circuit
board through a reflow cycle. The unequal-length form shows the design
lever: lengthening the stiffer, lower-expansion steel segment relative
to the copper lowers $F$, because the series compliance $L_s/E_s$ grows
and softens the restraint. A free (unclamped) bonded bar carries no
axial force but bows instead, the bimetallic-strip action a thermostat
exploits; the clamped case here converts the same mismatch into a pure
axial stress.

\subsection{The plane-strain crack-tip field and the $K$-dominant zone}

The fracture-mechanics mastery box states that the stress near a sharp
crack tip is singular and that its whole strength is carried by the
stress-intensity factor $K$. Carrying one numerical instance through the
near-tip field fixes the idea and prepares the third case study. A wide
steel plate under remote tension $\sigma = 120\,\text{MPa}$ contains a
through-thickness central crack of total length $2a = 24\,\si{\milli
\meter}$, so the half-length is $a = 12\,\si{\milli\meter}$. The steel
has fracture toughness $K_{IC} = 60\,\text{MPa}\sqrt{\text{m}}$. We want
the stress-intensity factor, the stress at a point $1\,\si{\milli\meter}$
directly ahead of the tip, and the size of the small plastic zone the
elastic singularity implies.

The stress-intensity factor for a central crack in a wide plate is $K =
\sigma\sqrt{\pi a}$ with the geometry factor $Y = 1$,
\[
K = \sigma\sqrt{\pi a} = (120)\sqrt{\pi(0.012)} = 120\sqrt{0.0377} =
120(0.1942) = 23.3\,\text{MPa}\sqrt{\text{m}}.
\]
Against the toughness $K_{IC} = 60\,\text{MPa}\sqrt{\text{m}}$ the crack
is stable: $K < K_{IC}$, with a margin of $60/23.3 = 2.6$ on the
stress-intensity factor (equivalently a margin of $2.6^2 = 6.6$ on the
applied stress at fixed crack size, the square again, as for the
flywheel speed). The near-tip stress directly ahead of the crack
($\phi = 0$) is $\sigma_{yy} = K/\sqrt{2\pi\rho}$, so at $\rho =
1\,\si{\milli\meter}$,
\[
\sigma_{yy} = \frac{K}{\sqrt{2\pi\rho}} = \frac{23.3\times 10^6}{\sqrt{2
\pi(0.001)}} = \frac{23.3\times 10^6}{0.0793} = 2.94\times 10^8\,\text{
Pa} \approx 294\,\text{MPa},
\]
already more than twice the remote $120\,\text{MPa}$ a millimetre from
the tip, and rising without bound as $\rho \to 0$. The unbounded
elastic stress cannot be real; the material yields in a small zone at
the tip. Equating the elastic $\sigma_{yy}$ to the yield stress $\sigma_y
= 350\,\text{MPa}$ gives a first estimate of the plastic-zone size,
\[
r_p = \frac{1}{2\pi}\left(\frac{K}{\sigma_y}\right)^2 = \frac{1}{2\pi}
\left(\frac{23.3}{350}\right)^2 = \frac{1}{2\pi}(0.00443) = 7.0\times
10^{-4}\,\text{m} \approx 0.7\,\si{\milli\meter}.
\]
The plastic zone is small compared to the crack length ($0.7$ against
$12\,\si{\milli\meter}$), which is the condition that makes $K$ the
governing parameter: a region surrounding the tip, larger than the
plastic zone but smaller than the crack, is dominated by the single
$K\!/\!\sqrt{2\pi\rho}$ term, and everything outside the tip detail
communicates with the crack only through the scalar $K$. This
\engterm{small-scale-yielding} condition is the licence for
linear-elastic fracture mechanics, and its failure (when the plastic
zone grows to the order of the crack or the remaining ligament) is the
boundary where elastic-plastic fracture mechanics and the $J$-integral
take over. The numbers here, $K = 23\,\text{MPa}\sqrt{\text{m}}$ against
a toughness of $60$, a sub-millimetre plastic zone in a centimetre-scale
crack, are the regime the next case study lives in
\cite{text:hertzberg-fatigue}.

\subsection{Traction on an inclined plane by Cauchy's formula}

The transformation formulas of section 3.5 give the stress on faces
rotated within the coordinate plane. Cauchy's formula does the more
general job: it returns the traction on a plane of \emph{any}
orientation, and reducing that traction to its normal and shear parts is
the calculation behind every question of the form ``what stress acts
across this particular weld, joint, or slip plane.'' A point carries the
stress tensor (in MPa)
\[
[\sigma] = \begin{pmatrix} 100 & 40 & 0 \\ 40 & 20 & 0 \\ 0 & 0 & -30
\end{pmatrix},
\]
and we want the traction on the plane whose unit normal is $\hat{n} = (
\tfrac{2}{3}, \tfrac{2}{3}, \tfrac{1}{3})$, then the normal stress and
the shear magnitude on that plane.

The traction vector is the matrix-vector product $\vec{t} = [\sigma]
\hat{n}$, computed row by row,
\[
t_x = 100(\tfrac{2}{3}) + 40(\tfrac{2}{3}) + 0 = \tfrac{280}{3} = 93.3,
\quad t_y = 40(\tfrac{2}{3}) + 20(\tfrac{2}{3}) + 0 = \tfrac{120}{3} =
40,
\]
\[
t_z = 0 + 0 + (-30)(\tfrac{1}{3}) = -10,
\]
all in MPa, so $\vec{t} = (93.3, 40, -10)\,\text{MPa}$. The traction is
not parallel to $\hat{n}$, which is the whole point: a general plane
carries both a normal push and a shear drag. The normal stress on the
plane is the projection of the traction onto the normal,
\[
\sigma_n = \vec{t}\cdot\hat{n} = 93.3(\tfrac{2}{3}) + 40(\tfrac{2}{3}) +
(-10)(\tfrac{1}{3}) = 62.2 + 26.7 - 3.3 = 85.6\,\text{MPa}.
\]
The shear magnitude is what Pythagoras leaves after the normal part is
removed from the traction,
\[
\tau_n = \sqrt{|\vec{t}|^2 - \sigma_n^2} = \sqrt{(93.3^2 + 40^2 + 10^2) -
85.6^2} = \sqrt{(8705 + 1600 + 100) - 7327} = \sqrt{3078} = 55.5\,\text{
MPa}.
\]
The plane carries $85.6\,\text{MPa}$ of normal stress and $55.5\,\text{
MPa}$ of shear. The method generalises the two-dimensional Mohr circle
to the full three-dimensional question and needs no rotation matrix: form
the traction, project it onto the normal for $\sigma_n$, and take the
Pythagorean remainder for $\tau_n$. The same three lines evaluate the
resolved shear stress on a crystallographic slip plane, the normal and
shear on a glued scarf joint, or the traction transmitted across any
internal surface an engineer chooses to interrogate.

\subsection{Axial deflection of a stepped bar by strain energy}

The strain-energy density of section 3.4 integrates over a member to give
its total stored energy, and the energy then yields deflections without
ever solving a displacement field, the method that becomes Castigliano's
theorem and the unit-load method in Volume III's later chapters. A
stepped steel tie rod hangs from a ceiling and carries an axial load $P =
60\,\text{kN}$ at its free end. The upper segment is $L_1 = 0.8\,\si{
\meter}$ long with area $A_1 = 600\,\si{\milli\meter}^2$; the lower
segment is $L_2 = 1.2\,\si{\meter}$ long with area $A_2 = 300\,\si{
\milli\meter}^2$; the steel has $E = 200\,\text{GPa}$. We want the total
elastic energy stored and the free-end extension, by energy.

The whole rod carries the same axial force $P$ (the load is axial and the
rod hangs in series), so each segment is in uniform tension with stress
$\sigma_i = P/A_i$ and uniform strain energy density $u_i = \sigma_i^2/
(2E)$. The energy in a segment is the density times its volume $A_i L_i$,
\[
U_i = \frac{\sigma_i^2}{2E}\,A_i L_i = \frac{(P/A_i)^2}{2E}\,A_i L_i =
\frac{P^2 L_i}{2 E A_i},
\]
the standard axial strain-energy term. Summing the two segments,
\[
U = \frac{P^2}{2E}\left(\frac{L_1}{A_1} + \frac{L_2}{A_2}\right) =
\frac{(60\times 10^3)^2}{2(200\times 10^9)}\left(\frac{0.8}{600\times
10^{-6}} + \frac{1.2}{300\times 10^{-6}}\right).
\]
The leading group is $(3.6\times 10^9)/(4.0\times 10^{11}) = 9.0\times
10^{-3}\,\text{J}\cdot\text{m}/\text{m}$, and the bracket is $1333 + 4000
= 5333\,\text{m}^{-1}$, so $U = (9.0\times 10^{-3})(5333) = 48\,\text{J}$.
Because the single applied load $P$ does work only through the deflection
$\delta$ at its own point, and the load builds linearly, the external
work is $\tfrac{1}{2}P\delta$, which must equal the stored energy. Solving
for the deflection,
\[
\delta = \frac{2U}{P} = \frac{2(48)}{60\times 10^3} = 1.6\times 10^{-3}\,
\si{\meter} = 1.6\,\si{\milli\meter}.
\]
The same number comes from summing the segment extensions $\delta =
\sum P L_i/(E A_i)$ directly, which is the algebraic identity behind the
energy route; the energy method earns its keep on structures where the
direct sum is hard, trusses and curved members and frames, while the
axial check here shows the bookkeeping is exact. The thinner lower
segment, half the area and longer, stores and stretches more: it
contributes $4000$ of the $5333$ in the bracket, three-quarters of the
deflection, the reminder that compliance pools in the slender, lightly
sectioned parts of a load path.

\subsection{Testing a strain field for compatibility}

Section 3.3 stated that a strain field must satisfy compatibility to come
from a continuous displacement. The test is a calculation, and running it
on a candidate field shows what compatibility rules out. A plane strain
field is proposed as $\epsilon_{xx} = c\,y^2$, $\epsilon_{yy} = c\,x^2$,
$\gamma_{xy} = 0$, with $c$ a constant. We check whether any continuous
displacement field produces it.

The two-dimensional compatibility equation requires
\[
\frac{\partial^2 \epsilon_{xx}}{\partial y^2} + \frac{\partial^2
\epsilon_{yy}}{\partial x^2} = \frac{\partial^2 \gamma_{xy}}{\partial x
\,\partial y}.
\]
The left side is $\partial^2(c y^2)/\partial y^2 + \partial^2(c x^2)/
\partial x^2 = 2c + 2c = 4c$. The right side is zero because $\gamma_{xy}
= 0$ everywhere. The equation reads $4c = 0$, satisfied only by the
trivial $c = 0$. A non-zero field of this form is therefore
incompatible: no continuous single-valued displacement produces it, and a
body asked to carry it would have to tear or overlap. The repair that
makes a near-relative compatible is instructive. Add a matching shear,
$\gamma_{xy} = 4c\,xy$; now the right side is $\partial^2(4c\,xy)/
(\partial x\,\partial y) = 4c$, the equation reads $4c = 4c$, and the
field is compatible. Integrating the strain-displacement relations
recovers a displacement $u_x = c\,x y^2 + f(x)$, $u_y = c\,x^2 y + g(y)$
consistent with all three strains up to a rigid-body motion, the
arbitrary functions absorbing translation and rotation. The lesson is the
one compatibility always teaches: the six strain components are not free
to be chosen independently, and a strain field handed to an engineer,
from a guessed Airy function or a coarse interpolation, must pass the
compatibility test before it can be believed to describe a real,
continuous body.

\subsection{A fully triaxial state: cubic roots and von Mises}

The principal-stress eigenproblem of section 3.5 reduces to a hand
formula only when one axis is already principal. A stress tensor with all
off-diagonal terms populated needs the full cubic, and carrying one
through fixes the three-dimensional machinery. A point carries (in MPa)
\[
[\sigma] = \begin{pmatrix} 60 & 20 & 20 \\ 20 & 40 & 10 \\ 20 & 10 & 20
\end{pmatrix},
\]
and we want the three invariants, the principal stresses, and the von
Mises equivalent, to check against $\sigma_y = 120\,\text{MPa}$.

The first invariant is the trace, $I_1 = 60 + 40 + 20 = 120\,\text{MPa}$.
The third is the determinant; expanding along the first row,
\[
I_3 = 60(40\cdot 20 - 10\cdot 10) - 20(20\cdot 20 - 10\cdot 20) + 20(20
\cdot 10 - 40\cdot 20),
\]
\[
= 60(800 - 100) - 20(400 - 200) + 20(200 - 800) = 60(700) - 20(200) +
20(-600) = 42000 - 4000 - 12000 = 26000.
\]
The second invariant is the sum of the principal $2\times 2$ minors,
\[
I_2 = (40\cdot 20 - 10^2) + (60\cdot 20 - 20^2) + (60\cdot 40 - 20^2) =
700 + 800 + 2000 = 3500.
\]
The characteristic cubic is $\sigma^3 - 120\sigma^2 + 3500\sigma - 26000
= 0$. Trying the divisors of the constant term, $\sigma = 20$ gives
$8000 - 48000 + 70000 - 26000 = 4000 \neq 0$, while $\sigma = 10$ gives
$1000 - 12000 + 35000 - 26000 = -2000$, so a root sits between $10$ and
$20$; a few bisection steps land it at $\sigma_3 \approx 11.6\,\text{MPa}$.
Dividing the cubic by $(\sigma - 11.6)$ leaves the quadratic $\sigma^2 -
108.4\,\sigma + 2241 \approx 0$, whose roots are $\sigma = 54.2 \pm
\sqrt{54.2^2 - 2241} = 54.2 \pm \sqrt{697} = 54.2 \pm 26.4$, that is
$80.6$ and $27.8\,\text{MPa}$. The three principal stresses are therefore
$\sigma_1 = 80.6$, $\sigma_2 = 27.8$, $\sigma_3 = 11.6\,\text{MPa}$, and a
check confirms the invariants: their sum is $120.0 = I_1$, and their
product is $(80.6)(27.8)(11.6) = 25{,}990 \approx I_3$, the small gap
being the rounding in the bisected root. The von Mises equivalent is
\[
\sigma_{\text{vM}} = \sqrt{\tfrac{1}{2}\big[(80.6 - 27.8)^2 + (27.8 -
11.6)^2 + (11.6 - 80.6)^2\big]} = \sqrt{\tfrac{1}{2}[2788 + 262 + 4761]}
= \sqrt{3906} = 62.5\,\text{MPa}.
\]
Against $\sigma_y = 120\,\text{MPa}$ the factor of safety is $120/62.5 =
1.9$, and the Tresca equivalent $\sigma_1 - \sigma_3 = 80.6 - 11.6 =
69.0\,\text{MPa}$ gives the slightly lower $120/69.0 = 1.7$, Tresca again
the more conservative. The example is the three-dimensional refrain: a
fully populated stress tensor reduces to three principal numbers through
its characteristic cubic, and the criterion is evaluated on those three,
exactly as the two-dimensional Mohr circle does for plane states, with
\texttt{code/yield\_criteria.py} returning the eigenvalues by machine for
any tensor handed to it.

\subsection{A crane hook: curved-beam stress with a shifted neutral axis}

Every bending example so far assumed a straight member, for which the
neutral axis passes through the centroid and the stress varies linearly
across the depth. A sharply curved member breaks both assumptions, and
the crane hook is the case where the difference decides the design. When
the radius of curvature is comparable to the section depth, the fibres on
the inside of the curve are shorter than those on the outside, so for a
given rotation of the section the inner fibres strain more, the neutral
axis shifts inward off the centroid, and the stress runs hyperbolically
rather than linearly across the depth. A hook of trapezoidal section
carries a working load $P = 50\,\text{kN}$ through its throat; the section
inner radius is $r_i = 40\,\si{\milli\meter}$, the outer radius $r_o =
100\,\si{\milli\meter}$, and for the arithmetic we approximate the
trapezoid by a rectangle of area $A = 1.8\times 10^{-3}\,\si{\meter}^2$
whose centroid sits at $\bar{r} = 70\,\si{\milli\meter}$. We want the
inner-fibre stress and the factor by which it exceeds the straight-beam
estimate.

The curved-beam result replaces the centroidal radius with the smaller
\engterm{neutral-axis radius} $R_n$, the radius at which the bending
strain, and so the bending stress, changes sign. For a rectangular
section of inner radius $r_i$ and outer radius $r_o$ it is the section
area divided by the integral of $dA/r$ across the depth, which for a
rectangle is
\[
R_n = \frac{r_o - r_i}{\ln(r_o/r_i)} = \frac{0.100 - 0.040}{\ln(0.100/
0.040)} = \frac{0.060}{\ln 2.5} = \frac{0.060}{0.916} = 0.0655\,\si{
\meter}.
\]
The neutral axis at $65.5\,\si{\milli\meter}$ sits inboard of the
centroid at $\bar{r} = 70\,\si{\milli\meter}$ by the eccentricity
$e = \bar{r} - R_n = 70.0 - 65.5 = 4.5\,\si{\milli\meter}$. The load
is an axial tension $P$ through the centroid plus a bending moment $M =
P\bar{r}$ about the centroidal axis, and the hoop stress at a fibre of
radius $r$ is
\[
\sigma_\theta(r) = \frac{P}{A} + \frac{M\,(R_n - r)}{A\,e\,r}, \qquad
e = \bar{r} - R_n = 4.5\,\si{\milli\meter},
\]
the small eccentricity that curved-beam theory magnifies into a large
inner-fibre stress.
The dominant term at the inner fibre is the bending part. With $M =
P\bar{r} = (50\times 10^3)(0.070) = 3500\,\text{N}\cdot\text{m}$, the
inner-fibre bending stress is $M(R_n - r_i)/(A\,e\,r_i) = (3500)(0.0655 -
0.040)/[(1.8\times 10^{-3})(4.5\times 10^{-3})(0.040)]$; the numerator is
$(3500)(0.0255) = 89.3$, the denominator is $3.24\times 10^{-7}$, giving
a bending contribution of $2.75\times 10^8 = 275\,\text{MPa}$ tensile.
Adding the direct term $P/A = (50\times 10^3)/(1.8\times
10^{-3}) = 27.8\,\text{MPa}$ gives an inner-fibre stress near $303\,\text{
MPa}$. Figure \ref{fig:vol03ch03:curved-beam-stress} draws the hyperbolic
profile against the straight-beam line. The straight-beam estimate would
have used $\sigma = P/A + M\,c/I$ with $c = 30\,\si{\milli\meter}$ and $I
= A(r_o - r_i)^2/12 = 5.4\times 10^{-7}\,\si{\meter}^4$, giving a bending
stress of $(3500)(0.030)/5.4\times 10^{-7} \approx 194\,\text{MPa}$ and a
total near $222\,\text{MPa}$. The curved-beam inner stress is thus about
$1.4$ times the straight-beam estimate, and the straight-beam formula is
unconservative at the inside of the curve, exactly the fibre where a hook
cracks. The lesson is geometric: once a member's radius of curvature
falls to the order of its depth, the neutral axis leaves the centroid and
the inner fibre carries a stress the straight-beam formula understates,
which is why hooks, chain links, C-frames, and press frames are sized by
curved-beam theory \cite[ch.~9]{text:crandall-mech-of-solids}.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-curved-beam-stress.tex}

\subsection{Hertzian contact: pressure and sub-surface shear in a bearing}

Every stress computed so far spread a load over a section the member
supplied. Two curved bodies pressed together supply almost no section at
all: they touch initially at a point or a line, and the contact patch
that carries the load grows only as the bodies deform, so the pressure in
the tiny patch climbs far above any nominal force-over-area. The elastic
contact problem Heinrich Hertz solved in 1882 sizes the patch and the
peak pressure. A steel ball of diameter $D = 12\,\si{\milli\meter}$
presses on a flat steel race under a load $F = 800\,\text{N}$; both bodies
are steel with $E = 200\,\text{GPa}$ and $\nu = 0.3$. We want the contact
patch radius, the peak contact pressure, and the sub-surface maximum
shear that governs rolling-contact fatigue.

For a sphere of radius $R = D/2 = 6\,\si{\milli\meter}$ on a flat, the
Hertz solution gives a circular contact patch of radius
\[
a = \left(\frac{3 F R}{4 E^*}\right)^{1/3},
\]
where $E^*$ is the \engterm{contact modulus} combining both bodies,
$1/E^* = (1-\nu_1^2)/E_1 + (1-\nu_2^2)/E_2$. For two identical steels
$1/E^* = 2(1 - 0.09)/(200\times 10^9) = 9.1\times 10^{-12}$, so $E^* =
1.10\times 10^{11}\,\text{Pa}$. The patch radius is
\[
a = \left(\frac{3(800)(0.006)}{4(1.10\times 10^{11})}\right)^{1/3} =
\left(\frac{14.4}{4.40\times 10^{11}}\right)^{1/3} = (3.27\times
10^{-11})^{1/3} = 3.20\times 10^{-4}\,\si{\meter},
\]
about $0.32\,\si{\milli\meter}$, a patch smaller than a pinhead carrying
the full $800\,\text{N}$. The peak pressure sits at the patch centre and
is one and a half times the mean pressure over the patch,
\[
p_0 = \frac{3F}{2\pi a^2} = \frac{3(800)}{2\pi(3.20\times 10^{-4})^2} =
\frac{2400}{6.43\times 10^{-7}} = 3.73\times 10^9\,\text{Pa} \approx
3730\,\text{MPa}.
\]
The peak contact pressure of $3.7\,\text{GPa}$ dwarfs the steel's yield
stress in a naive reading, and the resolution is the multiaxial state: at
the surface directly under the load the material sits in a nearly
hydrostatic compression, all three principal stresses large and negative
and close together, so the deviatoric part that drives von Mises yield is
small even though the pressure is enormous. Figure
\ref{fig:vol03ch03:hertz-contact} shows the semi-elliptical pressure
distribution and marks where the trouble actually lives. The maximum
shear stress is not at the surface but a short distance below it, at a
depth near $0.48a$, where the confinement relaxes and the deviatoric
stress peaks at about $\tau_{\max} \approx 0.31\,p_0 = 0.31(3730) =
1160\,\text{MPa}$. That sub-surface shear is what fatigues a bearing: a
crack initiates below the surface where the alternating shear is largest,
grows, and finally spalls a flake off the raceway, the pitting that ends a
bearing's life. The design levers follow from the exponents. The peak
pressure grows only as $F^{1/3}$, so contact stress is stubbornly
insensitive to load, while a larger ball radius $R$ or a conforming
(grooved) race enlarges the patch and drops $p_0$, which is why bearing
races are ground to conform closely to the ball and why raising the load
buys less relief than intuition expects \cite[ch.~3]{text:shigley2020}.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-hertz-contact.tex}

\subsection{Off-axis stress in a composite ply: transformation with two moduli}

The stress-transformation rule of section 3.5 rotated the stress tensor
alone, and for an isotropic material that is the whole job, because the
stiffness looks the same in every direction. A fibre-reinforced composite
ply is stiff and strong along its fibres and compliant across them, so a
load applied off the fibre axis must be resolved onto the fibre-aligned
axes before the material's direction-dependent strength can judge it. A
unidirectional carbon-epoxy ply carries an in-plane stress state
$\sigma_x = 120\,\text{MPa}$, $\sigma_y = -30\,\text{MPa}$, $\tau_{xy} =
40\,\text{MPa}$ in the laminate axes, and its fibres run at $\theta =
30^\circ$ to the $x$-axis. We resolve the stress onto the fibre-aligned
axes to compare against the ply's very different longitudinal, transverse,
and shear strengths.

The stress components transform by the same rule the isotropic case uses,
because the transformation of the stress tensor is a property of the
tensor and not of the material. Rotating the axes to the fibre direction
($1$ along the fibres, $2$ across them) by $\theta = 30^\circ$,
\[
\sigma_1 = \frac{\sigma_x + \sigma_y}{2} + \frac{\sigma_x - \sigma_y}{2}
\cos 2\theta + \tau_{xy}\sin 2\theta,
\]
with $(\sigma_x + \sigma_y)/2 = 45$, $(\sigma_x - \sigma_y)/2 = 75$,
$\cos 60^\circ = 0.5$, $\sin 60^\circ = 0.866$,
\[
\sigma_1 = 45 + 75(0.5) + 40(0.866) = 45 + 37.5 + 34.6 = 117.1\,\text{
MPa},
\]
the fibre-direction stress. The transverse stress is the complementary
combination,
\[
\sigma_2 = 45 - 37.5 - 34.6 = -27.1\,\text{MPa},
\]
and the in-plane shear on the fibre axes is
\[
\tau_{12} = -\frac{\sigma_x - \sigma_y}{2}\sin 2\theta + \tau_{xy}\cos
2\theta = -75(0.866) + 40(0.5) = -64.9 + 20 = -44.9\,\text{MPa}.
\]
The fibre-aligned state is $(\sigma_1, \sigma_2, \tau_{12}) = (117, -27,
-45)\,\text{MPa}$. The material check now runs component by component
against separate strengths, not against a single yield stress, because
the ply has no single strength. A typical carbon-epoxy ply carries a
longitudinal tensile strength near $1500\,\text{MPa}$, so $\sigma_1 =
117$ is comfortable at a margin above ten; a transverse strength of only
$50\,\text{MPa}$ in tension and perhaps $200\,\text{MPa}$ in compression,
so $\sigma_2 = -27$ compressive is safe with a margin near seven; and an
in-plane shear strength near $70\,\text{MPa}$, so $\tau_{12} = 45$ carries
a margin of about $1.6$, the governing check. The maximum-stress
criterion passes the ply, but the small shear margin is the warning a
Tsai-Wu interactive criterion would sharpen \cite[ch.~5]{text:chawla-composites}.
The lesson separates two ideas the isotropic chapter kept fused: the
stress transforms by geometry alone, but the strength it must be judged
against transforms with the material, and in an anisotropic ply the worst
component is rarely the largest one. The transverse and shear directions,
weak by construction, govern a composite long before the mighty fibre
direction is troubled, which is why laminates are stacked in several
orientations so that some ply always has fibres near any applied load.

\subsection{A thick spherical pressure vessel}

The thin-wall vessel and the Lam\'e cylinder both treated cylinders. A
sphere carries pressure more efficiently and fails differently, and the
thick-sphere solution is worth one pass because it is the geometry of a
diving sphere, a spherical gas accumulator, and the hemispherical head
that closes a cylindrical vessel. A thick steel sphere of inner radius
$a = 200\,\si{\milli\meter}$ and outer radius $b = 260\,\si{\milli\meter}$
holds an internal pressure $p = 40\,\text{MPa}$. We want the hoop stress
at the bore and at the outer surface, and a comparison to the thin-wall
estimate.

For a thick sphere under internal pressure the radial and tangential
stresses are, by the spherically symmetric analogue of the Lam\'e
solution,
\[
\sigma_r(r) = \frac{p\,a^3}{b^3 - a^3}\left(1 - \frac{b^3}{r^3}\right),
\quad \sigma_t(r) = \frac{p\,a^3}{b^3 - a^3}\left(1 + \frac{b^3}{2r^3}
\right),
\]
where the tangential stress $\sigma_t$ is the same in both directions
tangent to the sphere by symmetry. The geometry group is $a^3 = 8.0\times
10^{-3}\,\si{\meter}^3$, $b^3 = 1.758\times 10^{-2}\,\si{\meter}^3$, so
$b^3 - a^3 = 9.58\times 10^{-3}$ and $p\,a^3/(b^3 - a^3) = (40)(8.0\times
10^{-3})/(9.58\times 10^{-3}) = 33.4\,\text{MPa}$. At the bore $r = a$,
the ratio $b^3/a^3 = 2.20$, so the tangential stress is
\[
\sigma_t(a) = 33.4\left(1 + \frac{2.20}{2}\right) = 33.4(2.10) = 70.1\,
\text{MPa},
\]
and the radial stress there is $-p = -40\,\text{MPa}$ as the boundary
demands. At the outer surface $r = b$, $b^3/b^3 = 1$, so $\sigma_t(b) =
33.4(1.5) = 50.1\,\text{MPa}$ and $\sigma_r(b) = 0$. The hoop stress
falls from $70$ at the bore to $50$ at the outer surface, a gentler
gradient than the cylinder's because the sphere shares the load in two
tangential directions at once. The thin-wall sphere formula $\sigma =
pr/(2t)$ with $r = a$ and $t = b - a = 60\,\si{\milli\meter}$ gives
$(40)(0.200)/(2)(0.060) = 66.7\,\text{MPa}$, within five percent of the
exact bore value, so the thin-wall estimate is usable here even though
the wall is a quarter of the radius, because the sphere's biaxial sharing
keeps the through-wall variation mild. The comparison to a cylinder of the
same bore, wall, and pressure is the design point: a thin cylinder carries
a hoop stress $pr/t = 133\,\text{MPa}$, twice the sphere's $67$, so a
sphere holds a given pressure with half the membrane stress and is the
efficient shape wherever geometry allows, while a cylinder is chosen for
its packing and its ease of manufacture despite the penalty. The
hemispherical head on a cylindrical vessel exploits exactly this: the
head can be half the thickness of the shell it closes and still match its
stress \cite[ch.~10]{text:crandall-mech-of-solids}.

\subsection{Primary versus secondary stress: when a stress is self-limiting}

The thermal-stress and interference-fit examples produced real stresses
that add to the failure state, yet a pressure-vessel code would treat them
differently from the pressure hoop stress, and the distinction is one the
chapter's single yield check does not make. A \engterm{primary stress} is
one required for equilibrium with an applied load: it does not relax as
the material yields, so if it exceeds the material's capacity the part
collapses. A \engterm{secondary stress} arises from a geometric or thermal
constraint, a differential expansion or a discontinuity, and it is
\engterm{self-limiting}: once the material yields locally the constraint
is satisfied by a small plastic strain and the stress relaxes, so a
secondary stress that momentarily exceeds yield does not by itself cause
collapse. The thermal stress of the restrained rail is the clean case. A
rail heated $\Delta T = 60\,\text{K}$ between rigid abutments carries
$-E\alpha\,\Delta T = -144\,\text{MPa}$, computed earlier; suppose the
rise is $150\,\text{K}$, which would demand $-360\,\text{MPa}$, above the
$250\,\text{MPa}$ yield. The rail cannot carry $-360\,\text{MPa}$; it
yields in compression at $-250\,\text{MPa}$, accommodates the rest of the
suppressed thermal strain by a small plastic shortening, and the stress
is capped at yield rather than climbing to the elastic prediction. The
thermal stress is self-limiting because it was driven by a fixed strain,
not a fixed force, and once yield supplies the strain the driver is
spent. Contrast the pressure hoop stress in a vessel wall: that stress
must balance the pressure force whatever the material does, so if it
reaches yield the wall thins, the stress rises further on the thinned
section, and the vessel bursts in a runaway. The design consequence is
the tiered allowable of the pressure-vessel codes: a primary membrane
stress is held to about two-thirds of yield, while a secondary stress is
allowed up to three times that limit, near twice yield, precisely because
it will shake down to an elastic cycle after a few load applications
rather than accumulate. The classification is not a property of the number
of megapascals; it is a property of what drives the stress, a load or a
constraint, and the same $200\,\text{MPa}$ is dangerous as a primary
membrane stress and benign as a secondary thermal stress. An engineer who
lumps every stress into one yield check either over-builds against
harmless thermal stress or, worse, under-builds by treating a load-driven
primary stress as though it would obligingly relax \cite[ch.~13]{text:dowling-mechanical-behavior}.

\begin{mastery}
\textbf{The Lam\'e thick-walled-cylinder solution.} The thin-wall hoop
formula $\sigma_\theta = pr/t$ of section 3.5 assumes the wall is thin
enough that the stress is uniform across it. When the wall is thick, the
hoop stress varies strongly through the thickness, and the governing
result is the Lam\'e solution \cite[ch.~10]{text:crandall-mech-of-solids}.
For an axisymmetric cylinder the only nontrivial equilibrium equation is
\[
\frac{d\sigma_r}{dr} + \frac{\sigma_r - \sigma_\theta}{r} = 0,
\]
which, combined with the plane-strain (or plane-stress) compatibility of
a radial displacement $u(r)$ and Hooke's law, reduces to the Euler
equation $r^2 u'' + r u' - u = 0$ with solution $u = C_1 r + C_2/r$.
Differentiating gives the radial and hoop stresses in the two-constant
form
\[
\sigma_r(r) = A - \frac{B}{r^2}, \quad \sigma_\theta(r) = A +
\frac{B}{r^2},
\]
the signature of the Lam\'e field: the two stresses share a constant part
$A$ and carry equal-and-opposite $1/r^2$ parts, so their sum
$\sigma_r + \sigma_\theta = 2A$ is constant through the wall. Figure
\ref{fig:vol03ch03:lame-thick-cylinder} plots both stresses across a wall
whose outer radius is twice its bore, showing the hoop stress peaking at
the bore and the radial stress running from $-p$ there to zero at the free
outer surface. The
constants $A$ and $B$ are fixed by the pressures on the two surfaces. For
internal pressure $p_i$ at radius $a$ and zero external pressure at
radius $b$,
\[
A = \frac{p_i a^2}{b^2 - a^2}, \quad B = \frac{p_i a^2 b^2}{b^2 - a^2},
\]
so the hoop stress at the bore is $\sigma_\theta(a) = p_i(b^2 + a^2)/(b^2
- a^2)$, the formula used in the interference-fit example, and it always
exceeds the internal pressure: a thick cylinder cannot be made
arbitrarily strong by adding wall, because the bore hoop stress
approaches $2A \to$ a finite limit as $b \to \infty$. The thin-wall
$pr/t$ is the leading term of this exact result when $b - a \ll a$. The
divergence of the two stresses at the bore is also why autofrettage
helps: pre-yielding the bore in manufacture leaves a residual compressive
hoop stress there that the service pressure must first overcome, raising
the pressure the cylinder survives. Residual stress used on purpose is
the subject the next mastery box would open; here the point is that the
thick-cylinder field is a two-constant solution of one equilibrium
equation, and the thin-wall formula is its first-order shadow.
\end{mastery}

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-lame-thick-cylinder.tex}

\begin{mastery}
\textbf{Linear-elastic fracture mechanics: the difference between $K_t$
and $K$.} The stress concentration factor $K_t$ of section 3.8 is a
\emph{dimensionless multiplier} on a nominal stress at a smooth
geometric feature: a hole, a fillet, a notch with a finite root radius.
A crack is a different object. At the tip of a sharp crack the elastic
stress field is not merely amplified by a finite factor; it is singular,
growing without bound as the inverse square root of the distance $\rho$
from the tip,
\[
\sigma_{ij}(\rho, \phi) = \frac{K}{\sqrt{2\pi\rho}}\, f_{ij}(\phi),
\]
where the angular functions $f_{ij}(\phi)$ are universal and the whole
strength of the singularity is carried by one scalar, the
\engterm{stress-intensity factor} $K$, with units of $\text{MPa}\sqrt{
\text{m}}$. George Irwin introduced $K$ in the late 1950s as the
load-side parameter of fracture, building on Griffith's 1920 energy
argument. For a through-crack of length $2a$ in a wide plate under
remote stress $\sigma$, $K = \sigma\sqrt{\pi a}$; geometry corrections
enter as a dimensionless factor $Y$ of order one, $K = Y\sigma\sqrt{\pi
a}$. Fast fracture occurs when $K$ reaches the material's
\engterm{fracture toughness} $K_{IC}$, a measured material property
(roughly $1\,\text{MPa}\sqrt{\text{m}}$ for window glass, $20$ to $40\,
\text{MPa}\sqrt{\text{m}}$ for aluminium alloys, $50$ to $200\,\text{
MPa}\sqrt{\text{m}}$ for structural steels and tougher alloys). The
distinction is operational, not pedantic: $K_t$ tells you the peak
stress at a rounded feature and is the right tool when no crack is
present; $K$ tells you whether an \emph{existing} crack will run and is
the right tool once a crack, weld defect, or fatigue initiation site
exists. The two are linked at the limit. As a notch root radius shrinks
toward zero the elastic $K_t$ diverges, signalling that the finite-stress
picture has broken down and the crack-tip picture must take over.
Fatigue crack growth then proceeds by the Paris law $da/dN = C(\Delta
K)^m$, with the stress-intensity \emph{range} $\Delta K$ over a cycle as
the driver and the exponent $m$ running from about $2$ to $4$ for
ductile metals \cite{paper:paris-erdogan1963}. The fracture-mechanics
machinery of crack-driven failure is the subject of Volume X; the
chapter-3 point is the boundary between the two regimes, finite
concentration and singular crack, and the recognition that a part with a
crack must be assessed on $K$, not on $K_t$
\cite{text:hertzberg-fatigue}.
\end{mastery}

\begin{mastery}
\textbf{Weibull weakest-link statistics for brittle strength.} A ductile
metal has a yield stress that is, to good engineering precision, a single
number: two tensile coupons of the same alloy yield within a few percent
of each other. A brittle material does not. Two nominally identical
alumina rods can break at stresses that differ by a factor of two,
because each breaks at its largest flaw and the largest flaw is a random
draw from the population of flaws the rod happens to contain. Strength of
a brittle material is therefore a statistical quantity, and the working
description is the distribution Waloddi Weibull proposed in 1939
\cite{paper:weibull1939}. The argument is the weakest link. Divide the
stressed volume into many small independent elements; the part survives a
stress $\sigma$ only if every element survives it, so the survival
probability of the whole is the product of the survival probabilities of
the parts. If a single element of volume $V_0$ survives with probability
$\exp[-(\sigma/\sigma_0)^m]$, a body of volume $V$ made of $V/V_0$
independent elements survives with probability
\[
P_s(\sigma) = \exp\!\left[-\frac{V}{V_0}\left(\frac{\sigma}{\sigma_0}
\right)^{m}\right],
\]
the two-parameter Weibull form, with $\sigma_0$ a reference (scale)
strength and $m$ the \engterm{Weibull modulus} that sets the scatter. A
large $m$ (say $20$ and up, a tightly controlled material) means little
scatter and a strength approaching a single number; a small $m$ (say $5$
to $10$, ordinary ceramics) means wide scatter and a strength that must
be quoted as a distribution. Two consequences are the daily currency of
brittle design. First, the size effect: doubling the stressed volume
lowers the strength at fixed survival probability by the factor
$2^{-1/m}$, because a larger body offers more chances to contain a fatal
flaw, so a large ceramic part is weaker than a small coupon of the same
material and the laboratory strength must be volume-corrected before it
is used in a full-size component. Second, the design point: a brittle
part is not designed to a mean strength but to a low survival fraction,
commonly the stress at which $P_s = 0.99$ or higher, read off the lower
tail where the small-$m$ distribution falls away slowly and unforgivingly.
The contrast with the deterministic yield criterion of section 3.6 is the
heart of the matter. For a ductile metal the question is whether the von
Mises stress reaches a fixed $\sigma_y$; for a brittle ceramic the
question is the probability that the tensile stress reaches the largest
flaw the part happens to hold, and the answer is a number between zero and
one, not a yes or a no. The fracture-mechanics link closes the loop: each
flaw has its own critical stress $\sigma_c = K_{IC}/(Y\sqrt{\pi a})$ from
the LEFM box above, so the distribution of strengths is the distribution
of largest flaw sizes pushed through the $1/\sqrt{a}$ map, and a process
that tightens the flaw-size distribution (better powder, hot pressing,
proof testing to cull the weak tail) raises both $\sigma_0$ and $m$ at
once.
\end{mastery}

\begin{mastery}
\textbf{Residual stress: the stress a part carries with no load applied.}
Every result so far computed the stress a load produces. A part can also
carry stress with nothing applied at all, locked in during manufacture,
and this \engterm{residual stress} adds to the service stress and changes
the failure verdict. The mechanism is incompatible permanent strain. If
one region of a body is plastically stretched or thermally contracted by
a different amount than its neighbours, the regions no longer fit together
unstrained, and they settle into a self-equilibrating stress field that
satisfies equilibrium with zero external load: every tensile region is
balanced by a compressive region, so the section-integrated force and
moment vanish while the point-by-point stress does not. Welding is the
common source. The weld pool solidifies and then contracts as it cools,
but the cold surrounding metal restrains the contraction, so the weld and
its heat-affected zone end up in residual tension and the parent metal in
balancing compression, the tensile residual reaching the yield stress in
a typical as-welded joint. That is why the chapter's pressure-vessel case
carried weld residual stress into the upper end of its stress cycle, and
why post-weld heat treatment, which relaxes the locked-in strain by
letting the metal creep at temperature, is specified for fatigue- and
fracture-critical welds. Residual stress is not always the enemy. The
same physics, deliberately induced, is one of the most effective fatigue
defences in mechanical design. Shot peening hammers a thin surface layer
into compression; the autofrettage mentioned in the thick-cylinder box
pre-yields a gun-barrel or high-pressure-vessel bore to leave it in
residual compression; carburising and induction hardening leave the case
in compression by transformation and thermal mismatch. In each case the
surface, where fatigue cracks initiate, must first have its residual
compression overcome before the service stress can open a crack, so the
effective mean stress at the crack-prone surface falls and the fatigue
life rises, often several-fold. The accounting is the superposition of
section 3.1 run in reverse: the stress that matters for failure is the
sum of the applied stress and the residual stress, $\sigma_{\text{total}}
= \sigma_{\text{applied}} + \sigma_{\text{residual}}$, and a fatigue or
fracture check that omits the residual term is reading only half the
state. Measuring residual stress is correspondingly hard, since no
external load reveals it; X-ray and neutron diffraction read it from the
lattice spacing, and the hole-drilling method reads it from the strain
released when a small hole relaxes the locked-in field. The chapter's
single sentence on the subject is that a part's stress state is what it
carries, not only what is applied to it, and the difference is the
residual field built in before the part ever entered service.
\end{mastery}

\begin{mastery}
\textbf{Plane stress and plane strain: the two idealisations of a
two-dimensional problem.} The chapter's Mohr circle and its worked plane
states all treated three components, $\sigma_{xx}$, $\sigma_{yy}$,
$\sigma_{xy}$, and set the third dimension aside, but there are two
distinct ways to set it aside and they are not interchangeable.
\engterm{Plane stress} is the state of a thin plate loaded in its own
plane: the two faces are free, so the out-of-plane stress vanishes,
$\sigma_{zz} = \tau_{xz} = \tau_{yz} = 0$, while the out-of-plane strain
$\epsilon_{zz}$ is nonzero, the plate free to thin or thicken through its
thickness by the Poisson response to the in-plane stresses,
$\epsilon_{zz} = -\nu(\sigma_{xx} + \sigma_{yy})/E$. \engterm{Plane
strain} is the opposite idealisation, the state deep inside a long
prismatic body loaded transversely and uniform along its length, a dam, a
tunnel lining, a long roller, or the mid-thickness of a thick plate: the
material cannot move along the axis because its neighbours hold it, so the
out-of-plane strain vanishes, $\epsilon_{zz} = \gamma_{xz} = \gamma_{yz} =
0$, while the out-of-plane stress is nonzero and follows from the
constraint, $\sigma_{zz} = \nu(\sigma_{xx} + \sigma_{yy})$. Each state
zeroes one out-of-plane quantity and lets the conjugate one float, and the
choice is dictated by the geometry, thin-in-$z$ for plane stress,
long-in-$z$ for plane strain. The distinction is not cosmetic. The
in-plane stress-strain law carries a different effective modulus in the
two cases: plane strain replaces $E$ by $E/(1-\nu^2)$ and $\nu$ by
$\nu/(1-\nu)$ in the plane Hooke's law, so a plane-strain body is
stiffer and develops a third principal stress the plane-stress body does
not. The most consequential place this matters is fracture toughness. The
same steel has a lower measured toughness in plane strain than in plane
stress, because the plane-strain constraint at a crack tip in a thick
section builds up the triaxial tension that suppresses the crack-tip
plasticity and lets the crack run at a lower load; this is why fracture
toughness is quoted as the plane-strain value $K_{IC}$, the conservative
lower bound reached once the section is thick enough to enforce plane
strain at the tip, and why a thin sheet of the same alloy tolerates a
longer crack than a thick plate. The rosette and combined-loading examples
of this chapter were plane-stress states, free surfaces with $\sigma_{zz}
= 0$; the deep-forging inversion example and the crack-tip field were
nearer plane strain. Naming which idealisation a two-dimensional problem
sits in is the move that fixes the third principal stress, and the third
principal stress is what the yield criterion and the toughness both read.
\end{mastery}

\section{An end-to-end case study: a lifting padeye}\pathtag{standard}

The chapter's apparatus pays off when an engineer must clear a real
part for service from loads alone. We take a lifting padeye, the flat
steel plate with a pin hole that rigging shackles attach to when a
crane lifts a load, and carry it from the lift load through nominal
stress, stress concentration, peak stress, a duty cycle of repeated
lifts, a fatigue check, and a verdict. The padeye is a good vehicle
because every move of the chapter appears in it once: axial and bearing
stress, the net section, the stress-concentration factor of a hole, the
reduction to mean and alternating stress under cyclic loading, and a
Goodman fatigue check ending in a number an engineer would sign.

\subsection{Geometry and loads}

The padeye is a steel plate of thickness $t = 20\,\si{\milli\meter}$,
width $w = 110\,\si{\milli\meter}$ at the section through the pin hole,
with a pin hole of diameter $d_h = 35\,\si{\milli\meter}$, welded at its
base to the lifted structure. Figure
\ref{fig:vol03ch03:padeye-geometry} shows the geometry, the net section
through the hole, and the two worst-stress points at the hole edge on
the diameter perpendicular to the load. The crane lifts a working load
of $W = 100\,\text{kN}$. We apply a dynamic amplification factor of
$1.2$ for the snatch and swing of a real lift, so the design peak load
the padeye sees is $P = 1.2 W = 120\,\text{kN}$. The plate is a
structural steel with yield $\sigma_y = 350\,\text{MPa}$, ultimate
$\sigma_u = 500\,\text{MPa}$, and a fully-corrected fatigue endurance
limit $S_e = 180\,\text{MPa}$ at the relevant life.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-padeye-geometry.tex}

\subsection{Nominal stresses}

Two nominal stresses matter. The first is the tensile stress on the net
section, the reduced cross-section the tension must cross at the hole.
The net area is the gross width minus the hole, times the thickness,
$A_{\text{net}} = (w - d_h)\,t = (0.110 - 0.035)(0.020) = (0.075)(0.020)
= 1.5\times 10^{-3}\,\si{\meter}^2 = 1500\,\si{\milli\meter}^2$. The net-
section tensile stress is
\[
\sigma_{\text{net}} = \frac{P}{A_{\text{net}}} = \frac{120\times 10^3}
{1.5\times 10^{-3}} = 8.0\times 10^7\,\text{Pa} = 80\,\text{MPa}.
\]
The second is the bearing stress where the pin presses on the hole, the
contact force spread over the projected contact area $d_h\,t$,
\[
\sigma_{\text{br}} = \frac{P}{d_h\,t} = \frac{120\times 10^3}{(0.035)
(0.020)} = \frac{120\times 10^3}{7.0\times 10^{-4}} = 1.71\times 10^8\,
\text{Pa} = 171\,\text{MPa}.
\]
Bearing checks against a higher allowable than tension because the
material is confined, but $171\,\text{MPa}$ is well under the bearing
allowable for this steel (commonly taken near $1.5\,\sigma_y \approx
525\,\text{MPa}$), so bearing does not govern. The tensile net section,
with its stress concentration still to be applied, is the controlling
location.

\subsection{Stress concentration and peak stress}

The hole concentrates the net-section stress at the two points on the
load-perpendicular diameter, the points marked in figure
\ref{fig:vol03ch03:padeye-geometry}. For a hole in a finite-width strip
the concentration factor is referred to the net-section stress and
depends on the ratio of hole diameter to width. Here $d_h/w = 35/110 =
0.32$. From the standard finite-width pin-hole charts
\cite[ch.~3]{text:shigley2020} a ratio near $0.3$ gives a
net-section concentration factor of about $K_t = 2.4$ (the value falls
from the infinite-plate $3.0$ as the hole grows relative to the width,
because the net section itself is already reduced). The peak stress at
the hole edge is
\[
\sigma_{\max} = K_t\,\sigma_{\text{net}} = 2.4 \times 80 = 192\,\text{
MPa}.
\]
Against the yield stress $\sigma_y = 350\,\text{MPa}$ the static factor
of safety on first yield at the peak point is $350/192 = 1.8$. For a
ductile steel under a single static lift, local yielding would
redistribute the peak and the part would not fail; the static check
passes with margin. The static verdict alone, though, is the wrong
question for a part that lifts repeatedly.

\subsection{Duty cycle, mean and alternating stress}

A lifting padeye does not see one static load; it sees a load that
rises from near zero (the hook attached, no lift) to the full lift and
back, once per lift, for the life of the appliance. The cyclic stress
at the worst point swings between a minimum and a maximum. Take the
minimum as the peak stress under the dead weight of the rigging and an
empty hook, which we estimate at ten percent of the full peak,
$\sigma_{\min} = 0.1\,\sigma_{\max} = 19\,\text{MPa}$. The maximum is
the full peak, $\sigma_{\max} = 192\,\text{MPa}$. The concentration is
already folded into these peak-point stresses, so the mean and
alternating components follow directly from them:
\[
\sigma_m = \frac{\sigma_{\max} + \sigma_{\min}}{2} = \frac{192 + 19}{2}
= 105\,\text{MPa}, \quad
\sigma_a = \frac{\sigma_{\max} - \sigma_{\min}}{2} = \frac{192 - 19}{2}
= 86\,\text{MPa}.
\]
A refinement separates the geometric concentration from the fatigue
notch sensitivity: the fatigue concentration $K_f$ is usually a little
below the elastic $K_t$ because real materials are not perfectly
notch-sensitive. Using the same $K_t = 2.4$ as a conservative $K_f$ and
applying it to the alternating part of the nominal net stress gives
operating components close to $(\sigma_m, \sigma_a) \approx (95, 86)\,
\text{MPa}$, the point we carry into the fatigue diagram. The mean
stress is positive and substantial; a positive mean stress is what
makes a fatigue check more than a comparison of the alternating stress
to the endurance limit.

\subsection{Fatigue check by the Goodman criterion}

The endurance limit $S_e = 180\,\text{MPa}$ is the alternating stress
the material survives indefinitely at \emph{zero} mean stress. A
positive mean stress lowers the alternating stress the part can take.
The modified-Goodman criterion captures the trade with a straight line
in the mean-alternating plane, from $S_e$ on the alternating axis to
the ultimate strength $S_u$ on the mean axis:
\[
\frac{\sigma_a}{S_e} + \frac{\sigma_m}{S_u} = \frac{1}{n_f},
\]
where $n_f$ is the fatigue factor of safety. With the operating point
$(\sigma_m, \sigma_a) = (95, 86)\,\text{MPa}$, $S_e = 180$, and $S_u =
500\,\text{MPa}$,
\[
\frac{1}{n_f} = \frac{86}{180} + \frac{95}{500} = 0.478 + 0.190 =
0.668, \quad n_f = \frac{1}{0.668} = 1.50.
\]
The Soderberg criterion replaces $S_u$ with the yield strength $S_y =
350\,\text{MPa}$, which is more conservative because it also guards
against first-cycle yield:
\[
\frac{1}{n_f^{\text{Sod}}} = \frac{86}{180} + \frac{95}{350} = 0.478 +
0.271 = 0.749, \quad n_f^{\text{Sod}} = 1.33.
\]

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-goodman-diagram.tex}

Figure \ref{fig:vol03ch03:goodman-diagram} plots both lines, the
first-cycle yield line $\sigma_m + \sigma_a = S_y$, and the operating
point. The point sits below all three lines, so the padeye survives at
infinite life by Goodman and by Soderberg, and does not yield on the
first cycle. The load line from the origin through the operating point
meets the Goodman line at $(142, 129)\,\text{MPa}$; the ratio of that
distance to the operating-point distance is the Goodman factor of
safety, $1.5$, the same number the algebra returned.

\subsection{Verdict}

The padeye passes. Static first yield at the peak point carries a
factor of safety of $1.8$; the Goodman fatigue check carries $1.50$ and
the more conservative Soderberg check $1.33$. For a lifting appliance, a
fatigue factor of safety near $1.5$ on a fully-corrected endurance limit
is on the lean side of standard rigging practice, which often asks for
a higher margin on lifting gear because the consequence of failure is a
dropped load. The honest engineering reading is that the part is
adequate as analysed but has little reserve, and three design moves
would buy margin cheaply: a larger pin hole reinforced with welded
cheek plates to drop the net-section stress, a higher-grade steel to
lift $S_e$, or a reamed and polished hole bore to raise the surface
finish factor inside $S_e$. The case study has carried one set of
numbers from a crane load to a signed fatigue verdict; the discipline,
not the particular numbers, is the transferable result. Every load-
bearing part clears service by some version of this same chain: loads,
nominal stress, concentration, peak stress, duty cycle, mean and
alternating components, a fatigue criterion, a factor of safety, a
verdict.

\section{A second end-to-end case study: a flywheel disk}\pathtag{standard}

The padeye case carried an externally-applied load through a fatigue
check. A spinning rotor poses a different problem, because the load is
not applied from outside at all: it is the rotor's own mass resisting
the centripetal acceleration the spin demands. A flywheel that stores
energy, a turbine disk, a grinding wheel, and a motor armature all share
this loading, and they share its danger. The stress rises with the
square of the rotational speed, so a rotor that is comfortable at its
rated speed can tear itself apart at a modest overspeed, and the
fragments leave at the rim velocity. We take a solid steel energy-storage
flywheel and carry it from speed through the internal stress field to a
burst margin and a verdict, then show what a central bore would cost.

\subsection{Geometry, material, and the loading that has no external force}

The flywheel is a solid steel disk of outer radius $b = 0.25\,\si{
\meter}$ and thickness $h = 40\,\si{\milli\meter}$, mounted on a shaft
and spun to a rated speed of $6000\,\text{rpm}$. The steel has density
$\rho = 7800\,\si{\kilogram\per\meter\cubed}$, Young's modulus $E =
200\,\text{GPa}$, Poisson's ratio $\nu = 0.3$, yield stress $\sigma_y =
350\,\text{MPa}$, and ultimate strength $\sigma_u = 600\,\text{MPa}$.
The angular speed is
\[
\omega = 6000 \times \frac{2\pi}{60} = 628\,\si{\radian\per\second},
\quad \omega^2 = 3.95\times 10^{5}\,\si{\per\second\squared}.
\]
Each ring of material at radius $r$ wants to fly outward and is held in
by the hoop tension of the material outside it; the inertial body force
per unit volume is $\rho\,\omega^2 r$, directed outward. That body force
plays the role the external load played for the padeye. The product
$\rho\,\omega^2 = (7800)(3.95\times 10^5) = 3.08\times 10^9\,\si{\pascal
\per\meter\squared}$ is the scale factor for every stress that follows.

\subsection{The stress field through the disk}

The closed-form elastic solution for a solid spinning disk of uniform
thickness gives the radial and hoop stresses as functions of radius
\cite[ch.~3]{text:shigley2020}:
\begin{align*}
\sigma_r(r) &= \frac{3+\nu}{8}\,\rho\,\omega^2\,(b^2 - r^2), \\
\sigma_\theta(r) &= \frac{\rho\,\omega^2}{8}\left[(3+\nu)\,b^2 -
(1+3\nu)\,r^2\right].
\end{align*}
Both stresses are largest at the centre, where they are equal (the disk
centre carries an equibiaxial stress, the same in every in-plane
direction), and the radial stress falls to zero at the free outer rim
while the hoop stress stays high to the edge. The centre stress is
\[
\sigma_{\text{ctr}} = \frac{3+\nu}{8}\,\rho\,\omega^2\,b^2 =
\frac{3.3}{8}(3.08\times 10^9)(0.0625) = 7.9\times 10^7\,\text{Pa}
\approx 79\,\text{MPa},
\]
and the rim hoop stress is
\[
\sigma_\theta(b) = \frac{1-\nu}{4}\,\rho\,\omega^2\,b^2 =
\frac{0.7}{4}(3.08\times 10^9)(0.0625) = 3.4\times 10^7\,\text{Pa}
\approx 34\,\text{MPa}.
\]
Figure \ref{fig:vol03ch03:rotating-disk-stresses} plots both curves
across the radius. The centre governs, and at the centre the stress
state is equibiaxial tension, $\sigma_1 = \sigma_2 = 79\,\text{MPa}$,
$\sigma_3 = 0$. For an equibiaxial state the von Mises equivalent equals
the in-plane value, $\sigma_{\text{vM}} = \sqrt{\sigma_1^2 -
\sigma_1\sigma_2 + \sigma_2^2} = \sigma_1 = 79\,\text{MPa}$, and the
Tresca equivalent is $\sigma_1 - \sigma_3 = 79\,\text{MPa}$ as well; the
two criteria agree at equibiaxial tension, as the yield-locus figure
\ref{fig:vol03ch03:yield-loci} shows they must.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-rotating-disk-stresses.tex}

\subsection{Burst margin}

At the rated $6000\,\text{rpm}$ the governing centre stress is
$79\,\text{MPa}$ against a yield stress of $350\,\text{MPa}$, a factor
of safety of $350/79 = 4.4$ on first yield. The number looks generous,
and for steady running it is. The danger lives in the square law. The
centre stress scales as $\omega^2$, so the speed at which the centre
first yields is
\[
\omega_y = \omega_{\text{rated}}\sqrt{\frac{\sigma_y}{\sigma_{\text{ctr}}}}
= 628\sqrt{\frac{350}{79}} = 628\sqrt{4.43} = 1322\,\si{\radian\per
\second}, \quad N_y \approx 12{,}600\,\text{rpm}.
\]
First yield arrives at about $12{,}600\,\text{rpm}$, a hair over twice
the rated speed; gross plastic flow and burst follow as the speed climbs
toward the ultimate-strength condition near $16{,}500\,\text{rpm}$.
A factor of safety of $4.4$ on stress is a factor of only $\sqrt{4.4}
\approx 2.1$ on speed, the conversion every rotor engineer keeps in
mind: a generous-looking stress margin is a thin speed margin, because
the stress climbs with the square of the speed. An overspeed trip on the
drive, set well below $12{,}600\,\text{rpm}$, is the engineered guard
against the square law, and the containment housing around any
high-energy rotor is the acknowledgement that the guard can fail.

\subsection{What a central bore would cost}

Mounting the disk on a shaft tempts the designer to bore a hole through
the centre. The bore sits exactly where the solid disk's stress peaks,
and a hole in a stress field concentrates it. For a disk with a small
central hole of radius $a \to 0$ the hoop stress at the bore is
\[
\sigma_\theta^{\text{bore}} = \frac{3+\nu}{4}\,\rho\,\omega^2\,b^2 =
2\,\sigma_{\text{ctr}},
\]
exactly twice the solid-disk centre stress, even for a vanishingly small
hole. At the rated speed the bore hoop stress is $2 \times 79 = 159\,
\text{MPa}$, and the yield-onset speed drops to
\[
N_y^{\text{bore}} = N_{\text{rated}}\sqrt{\frac{350}{159}} = 6000
\sqrt{2.20} \approx 8{,}900\,\text{rpm}.
\]
A pinhole bore has cut the speed margin from twice rated to about $1.5$
times rated, with no change in disk size, material, or rated speed. The
result is the chapter's stress-concentration lesson in its most
expensive form: the factor of two is independent of how small the hole
is, so even a clearance hole for a balancing pin pays the full penalty.
High-energy flywheels are therefore built solid and coupled to the shaft
by a face flange or a shrink-fitted hub at a larger radius, never by a
hole on the axis.

\subsection{Verdict}

The solid flywheel passes at its rated speed with a stress factor of
safety of $4.4$ and a speed margin to first yield of about $2.1$. The
design is sound provided the drive carries an overspeed trip set below
the $12{,}600\,\text{rpm}$ yield-onset speed and the rotor runs inside a
containment housing rated for a burst. The case has carried a rotor from
its speed alone, with no externally applied load, through the internal
stress field to a burst margin, and it has shown that a central bore
would halve the working stress margin. The transferable discipline is
the square law: for any rotating part, convert the stress margin to a
speed margin before declaring it safe, and treat the rated speed as a
floor below the real hazard, not as the hazard itself.

\section{A third end-to-end case study: fatigue-crack growth and
leak-before-break in a pressure vessel}\pathtag{standard}

The padeye cleared a part with a stress concentration but no crack; the
flywheel cleared a part on a static stress margin. A welded pressure
vessel poses the case neither covered: a part that already contains a
crack-like flaw at the moment it enters service, because welding leaves
flaws, and that must survive a known number of pressurisation cycles
without either bursting or, failing that, must leak in a way an operator
can detect before it bursts. This case carries a flaw from its
as-welded size, through the Paris-law integration of fatigue crack
growth, to a fracture check against the toughness and a leak-before-break
verdict. Every result of the two fracture-mechanics mastery boxes and the
crack-tip worked example appears here in a single chain, and the chain
ends in two numbers an inspector can act on: a re-inspection interval and
a leak margin.

\subsection{Vessel, flaw, and duty}

The vessel is a welded steel cylinder of inner radius $r = 0.9\,\si{
\meter}$ and wall thickness $B = 18\,\si{\milli\meter}$, cycled between
near-atmospheric and an operating pressure $p = 2.0\,\text{MPa}$ once per
operating cycle (a daily fill-and-empty, say). The steel has yield
$\sigma_y = 350\,\text{MPa}$, ultimate $\sigma_u = 520\,\text{MPa}$, and
plane-strain fracture toughness $K_{IC} = 60\,\text{MPa}\sqrt{\text{m}}$.
Weld inspection guarantees that no flaw deeper than $a_0 = 1.5\,\si{
\milli\meter}$ survives acceptance, so the design must assume a surface
flaw of exactly that depth sits at the worst location, the inner wall at
a longitudinal weld where the hoop stress is the membrane stress and
weld residual stress adds to it. Figure
\ref{fig:vol03ch03:edge-crack-geometry} shows the geometry: a
semi-elliptical surface flaw growing through the wall under the cyclic
hoop stress.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-edge-crack-geometry.tex}

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-crack-growth-life.tex}

\subsection{The driving stress and the stress-intensity range}

The hoop stress is the membrane stress that drives the crack. At
operating pressure $\sigma_{\theta} = pr/B = (2.0\times 10^6)(0.9)/
(0.018) = 1.0\times 10^8\,\text{Pa} = 100\,\text{MPa}$. The cycle runs
from the empty state to this maximum; taking the empty hoop stress at
ten percent of the full value gives a stress range $\Delta\sigma =
0.9\times 100 = 90\,\text{MPa}$ for the crack-growth driver and a maximum
$\sigma_{\max} \approx 110\,\text{MPa}$ once weld residual stress is
carried in the upper end (the residual field raises the mean but adds
little to the range). The stress-intensity range for a semi-elliptical
surface flaw is $\Delta K = Y\,\Delta\sigma\sqrt{\pi a}$ with a geometry
factor $Y \approx 1.12$ for a shallow surface crack. At the starting
depth,
\[
\Delta K_0 = (1.12)(90)\sqrt{\pi(0.0015)} = (1.12)(90)(0.0687) =
6.9\,\text{MPa}\sqrt{\text{m}},
\]
and at the through-wall depth $a = B = 18\,\si{\milli\meter}$ the range
has grown to $\Delta K_B = (1.12)(90)\sqrt{\pi(0.018)} = 24\,\text{MPa}
\sqrt{\text{m}}$. The range more than triples across the life because
$\Delta K$ scales with $\sqrt{a}$ and the crack deepens by a factor of
twelve.

\subsection{Integrating the Paris law to leak}

The crack grows by the Paris law $da/dN = C(\Delta K)^m$, with $C =
6.9\times 10^{-12}\,\text{m/cycle}$ and $m = 3$ for this ferritic steel
(metres per cycle, with $\Delta K$ in $\text{MPa}\sqrt{\text{m}}$)
\cite{paper:paris-erdogan1963}. Substituting $\Delta K = Y\Delta\sigma
\sqrt{\pi a}$ and separating variables,
\[
\frac{da}{a^{m/2}} = C\,(Y\Delta\sigma\sqrt{\pi})^{m}\,dN.
\]
For $m = 3$ the left side integrates to $-2\,a^{-1/2}$, so the cycle
count to grow the crack from $a_0$ to a depth $a$ is
\[
N = \frac{2}{C\,(Y\Delta\sigma\sqrt{\pi})^{3}}\left(\frac{1}{\sqrt{a_0}} -
\frac{1}{\sqrt{a}}\right).
\]
The leading group is $C(Y\Delta\sigma\sqrt{\pi})^3 = (6.9\times 10^{-12})
[(1.12)(90)(1.7725)]^3 = (6.9\times 10^{-12})(178.7)^3 = (6.9\times
10^{-12})(5.70\times 10^6) = 3.94\times 10^{-5}$. The starting and ending
inverse-root depths are $a_0^{-1/2} = (0.0015)^{-1/2} = 25.8\,\text{m}^
{-1/2}$ and $B^{-1/2} = (0.018)^{-1/2} = 7.45\,\text{m}^{-1/2}$. The
cycle count to grow the flaw through the full wall is therefore
\[
N_{\text{leak}} = \frac{2}{3.94\times 10^{-5}}(25.8 - 7.45) = (5.08\times
10^4)(18.35) \approx 9.3\times 10^{5}\,\text{cycles}.
\]
At one cycle per day this is several thousand years, and at a more
aggressive ten cycles per day it is several centuries: the as-welded
flaw, held
below $1.5\,\si{\milli\meter}$ by inspection, does not threaten the
vessel within any plausible service life. Figure
\ref{fig:vol03ch03:crack-growth-life} plots the depth against cycles. The
shape is the lesson: the crack spends about two thirds of its life below
$5\,\si{\milli\meter}$, where $\Delta K$ is small and growth is slow, and
covers the last two thirds of the wall in the final third of the life,
where $\Delta K$ and the cube on it have taken hold. Most of the safety
lives in the early, slow phase, and the inspection interval should be set
so that two inspections always bracket the slow phase, never the steep
one.

\subsection{The fracture check: does it leak or break?}

Growing through the wall produces a leak only if the crack has not first
reached the critical depth $a_c$ at which $K_{\max}$ equals the
toughness. The critical depth follows from $K_{IC} = Y\sigma_{\max}\sqrt{
\pi a_c}$, solved for $a_c$,
\[
a_c = \frac{1}{\pi}\left(\frac{K_{IC}}{Y\sigma_{\max}}\right)^2 =
\frac{1}{\pi}\left(\frac{60}{(1.12)(110)}\right)^2 = \frac{1}{\pi}(0.487)^2
= \frac{0.237}{\pi} = 0.075\,\text{m} = 75\,\si{\milli\meter}.
\]
The critical depth is $75\,\si{\milli\meter}$, far beyond the
$18\,\si{\milli\meter}$ wall. The crack therefore penetrates the wall and
leaks while it is still only a quarter of the depth that would cause fast
fracture, and the vessel announces its flaw by a detectable leak long
before it could burst. This is the \engterm{leak-before-break} condition,
and it is the design philosophy behind much of pressure-vessel and
pipeline safety: size the wall and choose the toughness so that the
through-wall crack length stays below the critical length, converting a
catastrophic burst into a manageable leak. The check is a competition
between two depths, the wall thickness and the critical crack depth, and
leak-before-break is the inequality $B < a_c$, here $18 < 75$, satisfied
with a wide margin. A lower-toughness steel flips the inequality: at
$K_{IC} = 25\,\text{MPa}\sqrt{\text{m}}$ the critical depth falls to
$a_c = (1/\pi)(25/123)^2 = 13\,\si{\milli\meter}$, below the wall, and
the same vessel would break before it leaked. Toughness, not strength,
buys leak-before-break, which is why pressure-vessel codes specify
minimum impact toughness at the lowest service temperature and why a
high-strength low-toughness steel is the wrong choice for a vessel even
though its yield stress flatters the static calculation.

\subsection{Verdict and the inspection interval}

The vessel passes, and it passes in the strong sense: the as-welded flaw
will not grow through the wall within any plausible service life, and
even if it somehow did, the vessel would leak rather than burst. The
transferable products of the analysis are two numbers. The first is the
leak-before-break margin, the ratio of critical depth to wall thickness,
$a_c/B = 75/18 = 4.2$, the factor by which the toughness exceeds what
leak-before-break requires. The second is the inspection interval. A
defensible rule sets the re-inspection so that a flaw just below the
detection threshold cannot grow to a dangerous fraction of the wall
before the next look; with the slow early growth shown in figure
\ref{fig:vol03ch03:crack-growth-life}, a flaw at the $1.5\,\si{\milli
\meter}$ detection limit takes more than $10^5$ cycles to reach even
$5\,\si{\milli\meter}$, so an interval of a few tens of thousands of
cycles carries a large margin. The case has carried a flaw from the
welding torch to a signed re-inspection interval, and it has shown that
the governing material property for a cracked part is toughness, the
property the static yield calculation never sees. The chain, flaw size,
stress range, stress-intensity range, Paris integration, critical depth,
leak-before-break, inspection interval, is the standard damage-tolerant
discipline, and Volume X develops it in full
\cite{text:hertzberg-fatigue}.

\section{An applied vignette: preload and fatigue in a bolted
joint}\pathtag{standard}

The three case studies cleared monolithic parts. A preloaded bolted
joint is a different animal, because the bolt and the clamped members
form a pair of springs in parallel, and the preload decides how an
external load shares between them. The non-obvious result, that a tightly
preloaded bolt feels only a small fraction of an external cyclic load, is
the reason engine head bolts and connecting-rod bolts survive billions of
cycles, and it follows from one stiffness ratio. We carry a single joint
from preload through an external duty cycle to a fatigue verdict, short
enough to read in one sitting and complete enough to sign.

A steel cylinder-head bolt of grade 10.9 ($\sigma_y = 830\,\text{MPa}$,
$\sigma_u = 1040\,\text{MPa}$) clamps a stack of members. The bolt is an
$M12$ with tensile-stress area $A_t = 84\,\si{\milli\meter}^2$ and an
effective stiffness $k_b = 0.4\times 10^9\,\text{N/m}$; the clamped
members are far stiffer, $k_m = 1.6\times 10^9\,\text{N/m}$. The joint is
assembled to a preload $F_i = 30\,\text{kN}$, and in service an external
tensile load on the joint cycles between zero and $P_{\text{ext}} =
12\,\text{kN}$ once per firing cycle. We want the fraction of the external
load the bolt actually feels, the bolt's mean and alternating stress, and
the Goodman fatigue margin, and we want to confirm the joint does not
gap.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-bolt-joint-diagram.tex}

The bolt and members share the external load in proportion to their
stiffnesses, the result figure \ref{fig:vol03ch03:bolt-joint-diagram}
draws. The \engterm{joint stiffness factor} is
\[
C = \frac{k_b}{k_b + k_m} = \frac{0.4}{0.4 + 1.6} = 0.20,
\]
so the bolt force rises by only $C\,P_{\text{ext}}$ when the external
load is applied, while the members shed $(1 - C)P_{\text{ext}}$ of their
clamp \cite[ch.~8]{text:shigley-mechanical-design}. The bolt force swings
between $F_i$ at zero external load and $F_i + C\,P_{\text{ext}}$ at full
load,
\[
F_{b,\min} = 30\,\text{kN}, \quad F_{b,\max} = 30 + (0.20)(12) = 32.4\,
\text{kN}.
\]
The $12\,\text{kN}$ external swing reaches the bolt as a $2.4\,\text{kN}$
swing, one fifth of it, because the stiff members absorb the other four
fifths by relaxing their compression. Dividing by the tensile-stress
area gives the bolt stresses,
\[
\sigma_{\min} = \frac{30\times 10^3}{84\times 10^{-6}} = 357\,\text{MPa},
\quad \sigma_{\max} = \frac{32.4\times 10^3}{84\times 10^{-6}} = 386\,
\text{MPa}.
\]
The mean and alternating components are
\[
\sigma_m = \frac{386 + 357}{2} = 371\,\text{MPa}, \quad \sigma_a =
\frac{386 - 357}{2} = 14\,\text{MPa}.
\]
The alternating stress is the fatigue driver, and it is tiny, $14\,\text{
MPa}$, even though the joint carries a substantial mean stress from
preload. That is the engineering point of preload: it trades a high,
steady mean stress, which fatigue tolerates well, for a small alternating
stress, which fatigue punishes. A loose bolt ($F_i \to 0$) would see the
full external load directly, with $C$ irrelevant once the members gap,
and the alternating stress would jump by more than an order of magnitude.

The fatigue check uses the modified-Goodman line with a thread-root
fatigue concentration. A rolled thread on a grade 10.9 bolt carries a
fully-corrected endurance limit near $S_e = 130\,\text{MPa}$ at the
thread root, with the root concentration already folded into the rated
value. The alternating stress at the root is the nominal alternating
stress raised by the fatigue factor; taking $K_f = 3.0$ for a standard
rolled thread, $\sigma_{a,\text{root}} = 3.0(14) = 42\,\text{MPa}$, while
the mean stays at the preload-dominated $\sigma_m = 371\,\text{MPa}$. The
Goodman fatigue factor of safety is
\[
\frac{1}{n_f} = \frac{\sigma_{a,\text{root}}}{S_e} + \frac{\sigma_m}{S_u}
= \frac{42}{130} + \frac{371}{1040} = 0.323 + 0.357 = 0.680, \quad n_f =
1.47.
\]
A separate check guards first-cycle yield at the root: the peak root
stress is $\sigma_m + \sigma_{a,\text{root}} = 371 + 42 = 413\,\text{MPa}$
against $\sigma_y = 830\,\text{MPa}$, a static margin of $2.0$. The last
check is separation: the members stay clamped as long as the external
load does not exhaust the preload, $P_{\text{ext}} < F_i/(1 - C) = 30/
0.80 = 37.5\,\text{kN}$, comfortably above the $12\,\text{kN}$ service
load, so the joint never gaps and the linear sharing the analysis assumed
holds throughout the cycle.

The joint passes, with a fatigue margin of $1.47$ and no risk of gapping.
The vignette compresses the chapter's combined-stress discipline into a
joint that two springs share: preload sets the mean, the stiffness ratio
$C$ sets how much of the external load reaches the bolt, the thread root's
$K_f$ raises the alternating part, and Goodman returns the verdict. The
transferable rule is that a bolted joint is a fatigue problem solved by
preload: tighten enough that the external load barely moves the bolt, and
the bolt's alternating stress, not its mean stress, decides its life.

\section{An applied vignette: variable-amplitude fatigue of a
wind-turbine tower flange bolt}\pathtag{standard}

The bolted-joint vignette carried a single alternating stress against an
endurance limit. A wind turbine does not oblige with a single amplitude.
The tower rocks under a wind that gusts, lulls, and turns, and a flange
bolt at the tower base sees a spectrum of stress ranges, from the small
swings of a calm afternoon to the large swings of a storm. The tool for a
spectrum is the S-N curve and a linear damage sum, and this vignette carries
one bolt from a load spectrum to a fatigue-life fraction an operator can
book.

A tower base flange is clamped by grade 8.8 studs. One stud carries a
mean tensile preload with a superimposed cyclic component set by the
overturning moment the wind puts on the tower. Structural monitoring bins
a representative year of the alternating stress at the stud into three
amplitude blocks: a low block $\sigma_{a,1} = 40\,\text{MPa}$ that recurs
$n_1 = 2.0\times 10^7$ times a year (ordinary turbulence), a middle block
$\sigma_{a,2} = 70\,\text{MPa}$ at $n_2 = 1.5\times 10^6$ times (gusts),
and a high block $\sigma_{a,3} = 110\,\text{MPa}$ at $n_3 = 4.0\times 10^4$
times (storms). The stud's S-N curve, corrected for the thread-root
concentration and surface finish, has an endurance limit $S_e =
60\,\text{MPa}$ below which cycles do no damage, and a finite-life branch
$\sigma_a = \sigma_f' (N)^{-1/k}$ with the fitted constants giving a life
$N_2 = 3.0\times 10^6$ cycles at $70\,\text{MPa}$ and $N_3 = 2.0\times
10^5$ cycles at $110\,\text{MPa}$. Figure \ref{fig:vol03ch03:sn-curve}
sketches the shape of such a curve, with the endurance limit as its flat
tail.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-sn-curve.tex}

The \engterm{Palmgren-Miner rule} sums the fractional damage each block
does. A block of $n_i$ cycles at an amplitude whose S-N life is $N_i$ uses
up the fraction $n_i/N_i$ of the fatigue life, and failure is predicted
when the running sum reaches one:
\[
D = \sum_i \frac{n_i}{N_i}, \qquad \text{failure at } D = 1.
\]
The low block sits below the endurance limit, $40 < 60\,\text{MPa}$, so it
does no damage and drops out, the first payoff of an endurance limit under
a spectrum: the vast majority of the cycles, the gentle everyday
turbulence, cost nothing. The middle and high blocks each spend a
fraction,
\[
\frac{n_2}{N_2} = \frac{1.5\times 10^6}{3.0\times 10^6} = 0.50, \qquad
\frac{n_3}{N_3} = \frac{4.0\times 10^4}{2.0\times 10^5} = 0.20,
\]
so the yearly damage is $D_{\text{year}} = 0.50 + 0.20 = 0.70$. At that
rate the flange bolt exhausts its fatigue life in $1/0.70 \approx 1.4$
years, far short of a turbine's twenty-year design life. The verdict is a
fail, and the vignette's value is in showing which block governs: the
storm block, rare at forty thousand cycles a year, is not the driver; the
gust block, at half the yearly damage, is. Raising the endurance limit so
that the $70\,\text{MPa}$ gust block also falls below it, by a larger
stud, a rolled rather than cut thread to lift $S_e$, or a higher preload
to shrink the alternating component through the stiffness ratio $C$ of the
previous vignette, would drop the yearly damage to the storm block alone,
$D_{\text{year}} = 0.20$, and a five-year life, still short but pointing at
the fix. The transferable discipline is that variable-amplitude fatigue is
a bookkeeping problem: bin the spectrum, read a life for each bin off the
S-N curve, sum the fractions, and let the sum, not the largest single
range, decide the life. The rule is linear and ignores the order of the
blocks, an approximation whose error a fuller sequence-sensitive analysis
would correct, but the linear sum is the standard first pass and it names
the governing block correctly here.

\section{An applied vignette: a silicon pressure-sensor
diaphragm}\pathtag{standard}

Every vignette so far worked in structural steel. A silicon
micromachined pressure sensor moves the same mechanics into a chip a few
millimetres across and turns stress into a voltage. A thin square silicon
diaphragm is etched into the sensor die; pressure on one face bends it,
the bending strains the silicon, and piezoresistors diffused at the
diaphragm edge convert the strain into a resistance change a bridge reads
out. The device is a plate-bending problem, a stress-concentration problem
at the diaphragm edge, and a strain-to-signal problem at once, and its
sizing is a direct application of the chapter's tools to a material with
$E = 170\,\text{GPa}$ and a fracture strength far above its working stress
but a brittleness that forbids yielding.

The diaphragm is a square membrane of side $2a = 1.0\,\si{\milli\meter}$
(half-side $a = 0.5\,\si{\milli\meter}$) and thickness $h = 20\,\si{\micro
\meter}$, clamped on all four edges into the thick silicon frame, and it
measures a full-scale pressure $p = 100\,\text{kPa}$. For a clamped square
plate the maximum bending stress sits at the middle of an edge and is
\[
\sigma_{\max} = \beta\,\frac{p\,a^2}{h^2},
\]
with $\beta \approx 0.31$ for a clamped square plate at the edge midpoint.
Putting in numbers, the aspect group is $p a^2/h^2 = (100\times 10^3)
(0.5\times 10^{-3})^2/(20\times 10^{-6})^2 = (100\times 10^3)(2.5\times
10^{-7})/(4.0\times 10^{-10}) = (100\times 10^3)(625) = 6.25\times 10^7\,
\text{Pa}$, so
\[
\sigma_{\max} = 0.31 \times 62.5\,\text{MPa} = 19.4\,\text{MPa}.
\]
Against silicon's fracture strength, commonly taken near $1\,\text{GPa}$
for a well-etched surface but as low as $100$ to $300\,\text{MPa}$ where
etch pits and edge damage raise a stress concentration, the working stress
of $19\,\text{MPa}$ carries a comfortable margin even against the flawed
lower bound, and a wide margin against the pristine value. The margin must
be wide, because silicon is brittle: it does not yield and redistribute a
local peak the way the chapter's ductile steels do, so a Weibull
weakest-link description governs its strength and the design point is a low
survival fraction on the flaw distribution, not a mean strength. The
diaphragm edge is where both the peak bending stress and the etch-induced
flaws live, which is exactly where the piezoresistors are placed to catch
the largest signal, so the sensor reads its output from the same location
its fracture risk concentrates.

The signal follows from the strain. The edge bending strain is
$\epsilon_{\max} = \sigma_{\max}/E = 19.4\times 10^6/170\times 10^9 =
1.14\times 10^{-4} = 114\,\mu\epsilon$ at full scale. A silicon
piezoresistor has a gauge factor near $100$, far above the $2$ of a metal
foil gauge, so the fractional resistance change at full scale is
$\Delta R/R = G_F\,\epsilon = 100(1.14\times 10^{-4}) = 1.14\times 10^{-2}$,
a bit over one percent, a large and easily conditioned signal from a
Wheatstone bridge. The chain is the chapter's rosette example run forward
and miniaturised: pressure sets the plate bending stress, the stress sets
the edge strain through Hooke's law, and the strain sets a resistance the
electronics read. The design levers are visible in the formula. Thinning
the diaphragm ($h$ down) or widening it ($a$ up) raises the stress and the
signal together as $a^2/h^2$, trading fracture margin for sensitivity, so
a low-pressure sensor uses a thin wide diaphragm for sensitivity and a
high-pressure sensor a thick narrow one for survival. The single relation
$\sigma_{\max} = \beta\,p a^2/h^2$ sizes the sensor across four decades of
pressure range, and the same brittleness caveat rides along at every
scale: design to a survival fraction, not to a yield stress silicon does
not have.

\section{An applied vignette: a Belleville washer as a
nonlinear spring}\pathtag{standard}

Every spring the chapter has touched, the axial rod, the helical coil of
the estimation exercise, stored energy along a straight load line and
returned a force proportional to deflection. A \engterm{Belleville
washer}, a shallow cone of steel disc that flattens under axial load, does
not: its force-deflection curve bends, plateaus, and can even fall, and
that nonlinearity is exactly why it holds a bolt preload steady while an
ordinary washer cannot. The device is a plate-bending problem whose stress
peaks at a single edge, and its sizing applies the chapter's tools to a
spring whose stiffness is a design variable rather than a fixed constant.

A coned-disc washer has an outer diameter $D_o = 40\,\si{\milli\meter}$,
inner diameter $D_i = 20\,\si{\milli\meter}$, thickness $t = 1.5\,\si{
\milli\meter}$, and free cone height $h = 1.2\,\si{\milli\meter}$
(measured as the rise of the cone before loading). The steel has $E =
200\,\text{GPa}$ and $\nu = 0.3$. We want the load to flatten the washer,
the peak stress at that load, and the property of the height-to-thickness
ratio that sets the shape of the curve.

The load-deflection response of a Belleville washer is governed by the
ratio $h/t$, here $1.2/1.5 = 0.8$. Figure
\ref{fig:vol03ch03:belleville-washer} sketches the family. A low ratio,
below about $0.4$, gives a nearly linear stiff spring; a ratio near $1.4$
flattens the curve into a near-constant force over a range of deflection,
the property that keeps a bolt tight as a gasket creeps; a ratio above
about $2.8$ produces a snap-through region of negative stiffness, the
over-centre action of a latch. Our $h/t = 0.8$ sits in the rising but
softening regime, stiff at first and gentler as it flattens. The load to
flatten the cone (deflection $s = h$) follows from the Almen-Laszlo plate
relation for a coned disc,
\[
F = \frac{4E}{1-\nu^2}\,\frac{t^3 h}{K_1 D_o^2}\,\frac{h}{t}\left(\frac{h}
{t}\cdot\frac{1}{2} + 1\right),
\]
a compact form in which $K_1$ is a dimensionless geometry factor near
$0.69$ for the diameter ratio $D_o/D_i = 2$. The stiff leading group is
$4E/(1-\nu^2) = 4(200\times 10^9)/0.91 = 8.79\times 10^{11}\,\text{Pa}$,
the cube-and-height factor is $t^3 h = (1.5\times 10^{-3})^3(1.2\times
10^{-3}) = 4.05\times 10^{-12}\,\si{\meter}^4$, the diameter group is
$K_1 D_o^2 = 0.69(0.040)^2 = 1.10\times 10^{-3}\,\si{\meter}^2$, and the
dimensionless bracket at $h/t = 0.8$ is $(0.8)(0.4 + 1) = 1.12$.
Assembling,
\[
F = (8.79\times 10^{11})\,\frac{4.05\times 10^{-12}}{1.10\times 10^{-3}}\,
(1.12) = (8.79\times 10^{11})(3.68\times 10^{-9})(1.12) \approx 3.6\,
\text{kN},
\]
so a single washer of this modest size carries several kilonewtons to
flatten, the high force from a short stroke that makes stacked Belleville
washers the standard way to store a large preload in a short axial length.
The peak stress is a bending stress concentrated at the inner top edge,
the corner marked in the figure, and for a flattened washer it runs near
\[
\sigma_{\max} \approx \frac{4E}{1-\nu^2}\,\frac{t\,h}{K_1 D_o^2}\cdot C
\approx 1.1\,\text{GPa},
\]
with $C$ a stress factor of order one; a stress near a gigapascal is why
Belleville washers are made from hardened spring steel with a yield above
$1400\,\text{MPa}$ and why the inner edge, not the average section, sizes
the washer. The engineering payoff is the constant-force behaviour. A
bolted joint clamped by a stack of washers at $h/t \approx 1.4$ holds its
preload almost unchanged as the gasket relaxes or the joint wears through a
few tenths of a millimetre, because the near-flat part of the curve
returns nearly the same force across that stroke, where a stiff plain
washer would shed most of its preload over the same relaxation. The
transferable idea is that a spring's stiffness is a shape a designer
chooses, not a constant the material hands over: coning a flat disc turns
a linear plate into a nonlinear spring whose curve can be tuned, by the one
ratio $h/t$, from stiff to constant-force to snap-through.

\input{volumes/03-force/ch03-stress-and-strain/figures/fig-belleville-washer.tex}

\section{Project}\pathtag{core}

\begin{project}[Instrument a cantilever beam; measure strain at
three locations; compare to theory]
The reader instruments a small wooden or aluminum cantilever beam
with strain gauges at three locations, loads the beam in known
increments, and compares the measured strains to the predictions of
linear elastic beam theory. The deliverable is a thirty- to
fifty-page engineering report.

\textbf{Track}. Physical instrumentation plus analysis.
\textbf{Hazard class}: low. Strain gauges and conditioning
electronics carry no significant hazard at the loads used. Bench
testing only.

\textbf{Apparatus}. A cantilever beam of $25\,\si{\milli\meter} \times
3\,\si{\milli\meter}$ aluminum or $30\,\si{\milli\meter} \times 5\,
\si{\milli\meter}$ wood (oak, maple, or similar), $300\,\si{\milli
\meter}$ free length, clamped at one end in a bench vice. Three foil
strain gauges of $120\,\si{\ohm}$ nominal resistance bonded to the
top surface at distances $50\,\si{\milli\meter}$, $150\,\si{
\milli\meter}$, and $250\,\si{\milli\meter}$ from the clamped end. A
quarter-bridge strain-gauge conditioner (a commercial unit or a
Wheatstone bridge built from a power supply and an instrumentation
amplifier). A multimeter or oscilloscope for the conditioner's
output. A small set of calibration weights ($50\,\si{\gram}$ to
$500\,\si{\gram}$).

\textbf{Procedure}. The reader bonds the strain gauges, builds or
calibrates the conditioner, applies the calibration weights at the
free end of the cantilever in $50\,\si{\gram}$ increments, and
records the conditioner's output at each load and each gauge
location. The reader also records the beam's tip deflection (with
a ruler or a dial indicator) at each load.

\textbf{Analysis}. The reader computes, from linear elastic beam
theory:

\begin{itemize}[leftmargin=*]
  \item The bending moment at each gauge location as a function of
    applied tip load.
  \item The predicted top-surface bending stress at each gauge
    location, $\sigma = M y_{\max} / I$, with $y_{\max} = h/2$ and
    $I = bh^3/12$.
  \item The predicted strain at each gauge location, $\epsilon =
    \sigma / E$.
  \item The predicted tip deflection $\delta = PL^3 / (3EI)$.
\end{itemize}

The reader plots measured strain versus predicted strain at each
gauge location and computes the linear-regression slope (should be
$1.00 \pm 0.05$ if the bonding, the gauge factor, the conditioner
gain, and the beam properties are all correct). The reader plots
measured tip deflection versus predicted tip deflection and computes
the slope similarly.

\textbf{Reflection ($1{,}500$ to $3{,}000$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
  \item Where the measured strain agrees with theory and where it
    departs; what caused the departure.
  \item Whether the strain at the gauge nearest the clamp is
    influenced by the boundary condition (Saint-Venant principle).
  \item The uncertainty of the elastic modulus $E$ used in the
    prediction; how the uncertainty propagates to the predicted
    strain.
  \item One quantity in the experiment whose measurement was the
    limiting source of error; how a second iteration would address
    it.
\end{itemize}

\textbf{Deliverable}. The bonded beam (photographed), the
conditioner schematic, the strain data, the deflection data, the
calculations, the plots, and the reflection.

\textbf{Time}. Six to twelve hours of setup (gauge bonding is the
slowest step), one to two hours of measurement, four to eight hours
of analysis and write-up.
\end{project}

\section{Exercises}

The exercises mix calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment. Solutions are
provided for calculation and derivation exercises; estimation
exercises receive published reference estimates with the editor's
working; design, diagnosis, failure-analysis, and judgment exercises
receive rubrics rather than single answers. Exercises are tagged
with their reader-path tier.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A steel rod of diameter $20\,\si{\milli\meter}$ carries an axial
tensile load of $50\,\text{kN}$. Find the normal stress in the rod.
State the result in MPa and as a fraction of the typical structural
steel yield stress $\sigma_y = 250\,\text{MPa}$.
\paragraph{Solution sketch.} Area $A = \pi d^2/4 = \pi(0.020)^2/4 =
3.14\times 10^{-4}\,\si{\meter}^2 = 314\,\si{\milli\meter}^2$. Stress
$\sigma = N/A = 50{,}000/3.14\times 10^{-4} = 159\,\text{MPa}$. As a
fraction of yield, $159/250 = 0.64$, so the rod runs at $64\,\%$ of
yield, a factor of safety of $1.57$ against first yield.
\end{exercise}

\begin{exercise}
A bar of mild steel ($E = 200\,\text{GPa}$, $\nu = 0.3$) has a
uniaxial stress $\sigma_{xx} = 100\,\text{MPa}$, $\sigma_{yy} =
\sigma_{zz} = 0$. Find the three engineering normal strains.
\paragraph{Solution sketch.} Uniaxial Hooke's law: $\epsilon_{xx} =
\sigma_{xx}/E = 100\times 10^6 / 200\times 10^9 = 5.0\times 10^{-4}$.
The transverse strains are $\epsilon_{yy} = \epsilon_{zz} = -\nu\,
\epsilon_{xx} = -0.3(5.0\times 10^{-4}) = -1.5\times 10^{-4}$ (lateral
contraction). The axial strain is $500\,\mu\epsilon$ in tension; each
transverse strain is $150\,\mu\epsilon$ in compression.
\end{exercise}

\begin{exercise}
A point in a body has stress state $\sigma_{xx} = 80\,\text{MPa}$,
$\sigma_{yy} = 40\,\text{MPa}$, $\sigma_{xy} = 30\,\text{MPa}$. Find
the two principal stresses, the orientation of the principal axes,
and the maximum shear stress.
\paragraph{Solution sketch.} This is the worked state of section 3.5.
Centre $= (80+40)/2 = 60$. Radius $R = \sqrt{((80-40)/2)^2 + 30^2} =
\sqrt{20^2 + 30^2} = \sqrt{1300} = 36.1\,\text{MPa}$. Principal stresses
$\sigma_{1,2} = 60 \pm 36.1 = 96.1$ and $23.9\,\text{MPa}$. Orientation
$\tan 2\theta_p = 2(30)/(80-40) = 1.5$, so $\theta_p = 28.2^{\circ}$.
Maximum in-plane shear $\tau_{\max} = R = 36.1\,\text{MPa}$, on planes
$45^{\circ}$ from the principal axes. See figure
\ref{fig:vol03ch03:mohr-circle}.
\end{exercise}

\begin{exercise}
For the stress state in the previous exercise, evaluate the von
Mises equivalent stress and the Tresca equivalent stress (using
$\sigma_3 = 0$ for the out-of-plane principal stress). Compare to a
yield stress of $\sigma_y = 100\,\text{MPa}$ and state whether the
material has yielded by either criterion.
\paragraph{Solution sketch.} From the previous exercise $\sigma_1 =
96.1$, $\sigma_2 = 23.9$, and take $\sigma_3 = 0$. Von Mises:
$\sigma_{\text{vM}} = \sqrt{\tfrac{1}{2}[(96.1-23.9)^2 + (23.9-0)^2 +
(0-96.1)^2]} = \sqrt{\tfrac{1}{2}[5213 + 571 + 9235]} = \sqrt{7510} =
86.7\,\text{MPa}$. Tresca: $\sigma_1 - \sigma_3 = 96.1\,\text{MPa}$.
Against $\sigma_y = 100$: von Mises says safe ($86.7 < 100$), Tresca
says safe but closer ($96.1 < 100$). Neither criterion predicts yield;
Tresca, as always, sits nearer the threshold.
\end{exercise}

\begin{exercise}
A thin-walled cylindrical pressure vessel of inner radius $r =
0.5\,\si{\meter}$ and wall thickness $t = 8\,\si{\milli\meter}$
contains gas at $p = 2\,\text{MPa}$. Find the hoop stress, the
longitudinal stress, and the maximum shear stress in the wall.
\paragraph{Solution sketch.} Hoop stress $\sigma_{\theta} = pr/t =
(2)(0.5)/(0.008) = 125\,\text{MPa}$. Longitudinal stress
$\sigma_z = pr/(2t) = 62.5\,\text{MPa}$, half the hoop. The radial
stress at the inner wall is $-p = -2\,\text{MPa}$, negligible against
the others (the thin-wall approximation drops it). The three principal
stresses are then $125$, $62.5$, and $\approx 0\,\text{MPa}$, so the
maximum shear is $\tau_{\max} = (\sigma_1 - \sigma_3)/2 = (125 - 0)/2 =
62.5\,\text{MPa}$. The hoop stress governs, which is why pressure
vessels split along a longitudinal seam.
\end{exercise}

\begin{exercise}
A circular shaft of diameter $40\,\si{\milli\meter}$ transmits a
torque of $500\,\text{N}\cdot\text{m}$. The maximum shear stress in
the shaft is $\tau_{\max} = 16T / (\pi d^3)$. Compute it. Compare
to a yield stress in shear of $\tau_y = \sigma_y/2$ for steel with
$\sigma_y = 350\,\text{MPa}$.
\paragraph{Solution sketch.} $\tau_{\max} = 16T/(\pi d^3) = 16(500)/
[\pi (0.040)^3] = 8000/(2.01\times 10^{-4}) = 39.8\,\text{MPa}$. The
Tresca shear yield is $\tau_y = \sigma_y/2 = 175\,\text{MPa}$. The
shaft runs at $39.8/175 = 0.23$ of shear yield, a factor of safety of
$4.4$, comfortable for a transmission shaft once fatigue and stress
concentration at keyways are added.
\end{exercise}

\begin{exercise}
A steel cable of cross-sectional area $200\,\si{\milli\meter}^2$
carries a tension of $50\,\text{kN}$. Find the normal stress and the
elongation of a $20\,\si{\meter}$ length of the cable. Take $E =
200\,\text{GPa}$.
\paragraph{Solution sketch.} Stress $\sigma = N/A = 50{,}000/200\times
10^{-6} = 250\,\text{MPa}$. Strain $\epsilon = \sigma/E = 250\times 10^6
/200\times 10^9 = 1.25\times 10^{-3}$. Elongation $\Delta L = \epsilon L
= (1.25\times 10^{-3})(20) = 0.025\,\si{\meter} = 25\,\si{\milli\meter}$.
A real stranded cable elongates more than this because the strands seat
and the effective modulus of a lay is below the wire's $E$; the
calculation gives the lower bound from the solid-wire idealisation.
\end{exercise}

\begin{exercise}
A circular hole of diameter $10\,\si{\milli\meter}$ pierces a wide
plate carrying remote uniaxial tension of $50\,\text{MPa}$. Find the
maximum stress at the hole's edge using the Kirsch solution.
Identify the two points on the hole's circumference where the
maximum occurs.
\paragraph{Solution sketch.} Kirsch gives $K_t = 3$ for a small
circular hole in a wide plate, so $\sigma_{\max} = 3\sigma_0 = 3(50) =
150\,\text{MPa}$. The factor is independent of hole diameter; the
$10\,\si{\milli\meter}$ value is irrelevant once the hole is small
relative to the plate. The maximum occurs at the two points on the
hole edge that lie on the diameter perpendicular to the load (the
``equator'' of the hole relative to the pull). At the two points on the
diameter parallel to the load, the stress is $-\sigma_0 = -50\,\text{MPa}$,
compressive.
\end{exercise}

\begin{exercise}
A standard $M16$ machine bolt of root diameter $13.5\,\si{\milli
\meter}$ carries an axial tensile load of $40\,\text{kN}$. Find the
nominal stress on the root. Apply a fatigue stress concentration of
$K_f = 3.0$ and state the effective fatigue stress.
\paragraph{Solution sketch.} Root area $A = \pi d_r^2/4 =
\pi(0.0135)^2/4 = 1.43\times 10^{-4}\,\si{\meter}^2 = 143\,\si{\milli
\meter}^2$. Nominal root stress $\sigma_{\text{nom}} = N/A = 40{,}000/
1.43\times 10^{-4} = 280\,\text{MPa}$. With the fatigue concentration,
the effective stress is $K_f\,\sigma_{\text{nom}} = 3.0(280) = 840\,
\text{MPa}$. This exceeds the endurance limit of ordinary steel by a
wide margin, which is why bolts in fatigue service are pre-tensioned to
keep the cyclic component small and are made from high-grade alloy.
\end{exercise}

\begin{exercise}
A point in a body has principal stresses $\sigma_1 = 200\,\text{MPa}$,
$\sigma_2 = 100\,\text{MPa}$, $\sigma_3 = -50\,\text{MPa}$. Evaluate
the von Mises and Tresca equivalent stresses. State which criterion
is more conservative.
\paragraph{Solution sketch.} Tresca: $\sigma_1 - \sigma_3 = 200 -
(-50) = 250\,\text{MPa}$. Von Mises: $\sigma_{\text{vM}} =
\sqrt{\tfrac{1}{2}[(200-100)^2 + (100-(-50))^2 + ((-50)-200)^2]} =
\sqrt{\tfrac{1}{2}[10{,}000 + 22{,}500 + 62{,}500]} = \sqrt{47{,}500} =
218\,\text{MPa}$. Tresca ($250$) exceeds von Mises ($218$), so Tresca
predicts yield at a lower applied load: it is the more conservative
criterion, as it is for every stress state.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the relations $G = E/[2(1+\nu)]$ and $K = E/[3(1-2\nu)]$ for
an isotropic linear elastic material. State the limiting values of
$\nu$ for which each result is finite and positive.
\paragraph{Solution sketch.} For $G$: apply a pure shear $\tau$ and
note that the principal stresses are $\sigma_1 = \tau$, $\sigma_2 =
-\tau$. The principal strain along axis $1$ is $\epsilon_1 = (1/E)
[\sigma_1 - \nu\sigma_2] = (\tau/E)(1+\nu)$. The engineering shear
strain at $45^{\circ}$ relates to the principal strains by $\gamma =
\epsilon_1 - \epsilon_2 = 2\epsilon_1 = 2\tau(1+\nu)/E$. Then $G =
\tau/\gamma = E/[2(1+\nu)]$. For $K$: apply hydrostatic stress
$\sigma_{xx} = \sigma_{yy} = \sigma_{zz} = -p$. Each normal strain is
$\epsilon = (1/E)[-p - \nu(-p-p)] = -p(1-2\nu)/E$. The volumetric
strain is $\epsilon_{kk} = 3\epsilon = -3p(1-2\nu)/E$, and $K =
-p/\epsilon_{kk} = E/[3(1-2\nu)]$. Limits: $G > 0$ requires $\nu > -1$;
$K > 0$ and finite requires $\nu < 1/2$. The physical range is
$-1 < \nu < 1/2$, with $\nu \to 1/2$ the incompressible limit ($K \to
\infty$).
\end{exercise}

\begin{exercise}
Show that, for a uniaxial stress state $\sigma_{xx} = \sigma$,
$\sigma_{yy} = \sigma_{xy} = 0$, the principal stresses are
$\sigma_1 = \sigma$, $\sigma_2 = 0$, and the maximum shear stress is
$\sigma/2$ at $45^{\circ}$ to the loading direction.
\paragraph{Solution sketch.} The matrix $\begin{psmallmatrix}\sigma &
0\\ 0 & 0\end{psmallmatrix}$ is already diagonal, so its eigenvalues
$\sigma_1 = \sigma$, $\sigma_2 = 0$ are the principal stresses and the
axes are already principal. The Mohr circle has centre $\sigma/2$ and
radius $\sqrt{(\sigma/2)^2 + 0} = \sigma/2$, so $\tau_{\max} = \sigma/2$
at the top of the circle, which corresponds to $2\theta = 90^{\circ}$,
i.e. $\theta = 45^{\circ}$ on the physical body. A bar pulled in
tension carries its largest shear on planes at $45^{\circ}$ to the
pull; ductile slip bands appear at exactly that angle.
\end{exercise}

\begin{exercise}
Derive Mohr's circle from the two-dimensional stress transformation
formulas in section 3.5. State the geometric construction explicitly.
\paragraph{Solution sketch.} Write the transformation as
$\sigma_{x'x'} - c = R\cos\phi$ and $\sigma_{x'y'} = -R\sin\phi$ with
$c = (\sigma_{xx}+\sigma_{yy})/2$, $R = \sqrt{((\sigma_{xx}-\sigma_{yy})
/2)^2 + \sigma_{xy}^2}$, and a phase $\phi = 2\theta - \alpha$ where
$\alpha$ is fixed by the original components. Squaring and adding
eliminates $\phi$: $(\sigma_{x'x'} - c)^2 + \sigma_{x'y'}^2 = R^2$. This
is the equation of a circle in the $(\sigma, \tau)$ plane of centre
$(c, 0)$ and radius $R$. The construction: plot $X = (\sigma_{xx},
\sigma_{xy})$ and $Y = (\sigma_{yy}, -\sigma_{xy})$, draw the circle on
$XY$ as diameter; the axis intercepts are the principal stresses, the
top of the circle is $\tau_{\max} = R$, and a physical rotation of
$\theta$ moves the diameter by $2\theta$ on the circle. The factor of
two between physical and circle angles is the geometric content of the
$\cos 2\theta$ in the transformation.
\end{exercise}

\begin{exercise}
For pure shear ($\sigma_{xx} = \sigma_{yy} = 0$, $\sigma_{xy} =
\tau$), show that the principal stresses are $\sigma_1 = \tau$,
$\sigma_2 = -\tau$, and that the von Mises equivalent stress is
$\sqrt{3} \tau$. Compare to the Tresca equivalent stress for the
same state.
\paragraph{Solution sketch.} The matrix $\begin{psmallmatrix}0 & \tau\\
\tau & 0\end{psmallmatrix}$ has eigenvalues $\pm\tau$, so $\sigma_1 =
\tau$, $\sigma_2 = -\tau$ (and $\sigma_3 = 0$). Von Mises:
$\sigma_{\text{vM}} = \sqrt{\tfrac{1}{2}[(\tau-(-\tau))^2 + (-\tau-0)^2
+ (0-\tau)^2]} = \sqrt{\tfrac{1}{2}[4\tau^2 + \tau^2 + \tau^2]} =
\sqrt{3\tau^2} = \sqrt{3}\,\tau$. Tresca: $\sigma_1 - \sigma_3 =
\tau - (-\tau) = 2\tau$. So von Mises predicts yield in pure shear at
$\tau = \sigma_y/\sqrt{3} \approx 0.577\sigma_y$, Tresca at $\tau =
\sigma_y/2 = 0.5\sigma_y$. The ratio of the two predicted shear yield
stresses is $2/\sqrt{3} \approx 1.155$, the maximum disagreement
between the criteria.
\end{exercise}

\begin{exercise}
Derive the thin-walled-cylinder hoop and axial stress formulas
$\sigma_{\text{hoop}} = pr/t$ and $\sigma_{\text{axial}} = pr/(2t)$
from equilibrium considerations on appropriate free-body diagrams of
half-cylinders. State the assumptions that make the formulas
accurate.
\paragraph{Solution sketch.} For the hoop stress, cut the cylinder
along a diametral plane through its length $\ell$. The pressure acts on
the projected area $2r\ell$, giving a bursting force $p(2r\ell)$. This
is resisted by the hoop stress on the two cut wall strips, each of area
$t\ell$, so $2\sigma_{\theta}(t\ell) = p(2r\ell)$, hence
$\sigma_{\theta} = pr/t$. For the axial stress, cut perpendicular to
the axis. The pressure acts on the end cap area $\pi r^2$, force
$p\pi r^2$, resisted by the axial stress on the annular wall of area
$2\pi r t$: $\sigma_z(2\pi r t) = p\pi r^2$, hence $\sigma_z = pr/(2t)$.
Assumptions: thin wall ($t \ll r$, so stress is uniform through the
thickness and the radial stress is negligible), closed ends or
remote-from-ends location, and no bending from supports.
\end{exercise}

\begin{exercise}[\textit{deferred to Chapter 8}]
Derive the bending stress formula $\sigma = -My/I$ for a slender
elastic beam under pure bending. State the assumptions used. Chapter
8 of this volume develops the full beam-theory machinery.
\paragraph{Solution sketch (outline; full treatment in Chapter 8).}
Assume plane sections remain plane and perpendicular to the neutral
axis (Euler-Bernoulli kinematics), linear elasticity, and small
deflection. Plane sections give a strain that varies linearly with the
distance $y$ from the neutral axis, $\epsilon = -y/\rho$ where $\rho$ is
the radius of curvature. Hooke's law gives $\sigma = E\epsilon =
-Ey/\rho$. Demanding that the stress distribution carry no net axial
force fixes the neutral axis at the centroid; demanding that it carry
moment $M$ gives $M = \int_A \sigma(-y)\,dA = (E/\rho)\int_A y^2\,dA =
EI/\rho$, so $1/\rho = M/(EI)$. Substituting back, $\sigma = -My/I$.
The chapter-8 development relaxes the pure-bending assumption to include
transverse shear.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the stress at the base of a $300\,\si{\meter}$ skyscraper's
central column, assuming the column carries the gravity load of all
the floors above it. State assumptions about the building's average
density and the column's cross-section.
\paragraph{Solution sketch.} Take a $300\,\si{\meter}$ tower of
footprint $50 \times 50\,\si{\meter} = 2500\,\si{\meter}^2$, $90$
storeys, average mass density of the occupied volume about $250\,\si{
\kilogram\per\meter\cubed}$ (structure plus contents, hollow). Total
mass $\approx (2500)(300)(250) = 1.9\times 10^8\,\si{\kilogram}$,
weight $W \approx 1.9\times 10^9\,\text{N}$. Suppose the vertical load
is carried by columns totalling $2\,\%$ of the floor area at the base,
$A \approx 50\,\si{\meter}^2$. Then $\sigma \approx W/A = 1.9\times 10^9
/50 = 38\,\text{MPa}$. Against structural steel yield ($250\,\text{MPa}$)
this is comfortable; the controlling design issue at this height is not
column crushing but lateral wind load and buckling. The estimate lands
in the right tens-of-MPa band, which is the point.
\end{exercise}

\begin{exercise}
Estimate the maximum bending stress in a $4\,\si{\meter}$ residential
floor joist of section $40\,\si{\milli\meter} \times 200\,\si{\milli
\meter}$ under a centred $80\,\si{\kilogram}$ person. Compare to the
typical allowable bending stress for structural wood ($\sim
15\,\text{MPa}$).
\paragraph{Solution sketch.} Treat the joist as simply supported, span
$L = 4\,\si{\meter}$, central point load $P = mg = 80(9.81) \approx
785\,\text{N}$. Maximum moment $M = PL/4 = 785(4)/4 = 785\,\text{N}
\cdot\text{m}$. Section: $b = 40\,\si{\milli\meter}$, $h = 200\,\si{
\milli\meter}$, so $I = bh^3/12 = (0.04)(0.2)^3/12 = 2.67\times 10^{-5}
\,\si{\meter}^4$ and $y_{\max} = h/2 = 0.1\,\si{\meter}$. Bending stress
$\sigma = My_{\max}/I = (785)(0.1)/2.67\times 10^{-5} = 2.9\,\text{MPa}$.
This is about a fifth of the $15\,\text{MPa}$ allowable, so one person
mid-span is no problem; the joist is sized by the distributed dead-plus-
live load of the whole floor and by deflection limits, not by a single
occupant.
\end{exercise}

\begin{exercise}
Estimate the strain in a strand of human hair holding a $50\,\si{\
gram}$ hanging mass. State the hair diameter and Young's modulus
assumed; bracket within a factor of three.
\paragraph{Solution sketch.} Hair diameter $d \approx 80\,\mu\text{m} =
8\times 10^{-5}\,\si{\meter}$, area $A = \pi d^2/4 \approx 5\times
10^{-9}\,\si{\meter}^2$. Load $W = 0.05(9.81) \approx 0.49\,\text{N}$.
Stress $\sigma = W/A \approx 0.49/5\times 10^{-9} \approx 1.0\times
10^8\,\text{Pa} = 100\,\text{MPa}$ (which is near hair's tensile
strength of $150$--$250\,\text{MPa}$, so a $50\,\si{\gram}$ mass on one
hair is close to breaking it). With Young's modulus of keratin $E
\approx 3\,\text{GPa}$, strain $\epsilon = \sigma/E \approx 1.0\times
10^8 / 3\times 10^9 \approx 0.03$, a few percent. Bracketing: a thinner
hair ($50\,\mu\text{m}$) or softer keratin ($2\,\text{GPa}$) pushes the
strain to $0.1$ or beyond, at which point the hair simply breaks. The
factor-of-three band straddles $0.01$ to $0.1$ strain.
\end{exercise}

\begin{exercise}
Estimate the maximum stress at the corner of a square window cutout
in a thin-walled pressure vessel of internal pressure $0.1\,\text{
MPa}$, wall thickness $2\,\si{\milli\meter}$, internal diameter
$2\,\si{\meter}$, and window cutout dimension $200\,\si{\milli
\meter}$. State assumptions about $K_t$ for the corner.
\paragraph{Solution sketch.} Nominal hoop stress on the wall (away from
the cutout): $\sigma_0 = pr/t = (0.1\times 10^6)(1.0)/(0.002) = 50\,
\text{MPa}$. A square cutout with a modest corner radius has an elastic
stress concentration of roughly $K_t \approx 3$ to $5$; take $K_t = 4$
for a lightly-rounded corner (a truly sharp corner sends $K_t$ much
higher). Peak corner stress $\sigma_{\max} \approx K_t\sigma_0 = 4(50) =
200\,\text{MPa}$. This is the de Havilland window-corner lesson in
miniature: a $50\,\text{MPa}$ nominal stress, comfortable for aluminium,
becomes a $200\,\text{MPa}$ peak at the corner, into the range where
cyclic pressurisation drives fatigue. Rounding the corner to drop $K_t$
toward $2$ halves the peak.
\end{exercise}

\begin{exercise}
Estimate the elastic energy stored in a steel spring of $300\,\si{
\milli\meter}$ rest length, $20\,\si{\milli\meter}$ outer diameter,
$2\,\si{\milli\meter}$ wire diameter, $30$ active turns, when
compressed by $40\,\si{\milli\meter}$. State assumptions about the
shear modulus and the spring constant; compare to the energy of
a $40\,\si{\milli\meter}$ free fall of a $1\,\si{\kilogram}$ mass.
\paragraph{Solution sketch.} Helical-spring rate $k = G d^4/(8 D^3 n)$
with wire diameter $d = 2\,\si{\milli\meter}$, mean coil diameter
$D \approx 18\,\si{\milli\meter}$ (outer $20$ minus one wire), $n = 30$
active turns, steel shear modulus $G = 80\,\text{GPa}$. Then $k =
(80\times 10^9)(2\times 10^{-3})^4 / [8(18\times 10^{-3})^3(30)] =
(80\times 10^9)(1.6\times 10^{-11}) / [8(5.83\times 10^{-6})(30)] =
1.28 / 1.40\times 10^{-3} \approx 915\,\text{N/m}$. Stored energy
$U = \tfrac{1}{2}k x^2 = \tfrac{1}{2}(915)(0.04)^2 = 0.73\,\text{J}$.
Compare to the free fall: a $1\,\si{\kilogram}$ mass dropping $40\,\si{
\milli\meter}$ releases $mgh = (1)(9.81)(0.04) = 0.39\,\text{J}$. The
spring stores about twice that, of the same order. The estimate is
sensitive to $D$ and $n$; a factor-of-two band is appropriate.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a Mohr's circle calculator that accepts $(\sigma_{xx},
\sigma_{yy}, \sigma_{xy})$ and returns $(\sigma_1, \sigma_2, \theta_p,
\tau_{\max})$. Verify on three textbook stress states.
\paragraph{Solution sketch.} The reference implementation is
\texttt{code/mohrs\_circle.py}: centre $= (\sigma_{xx}+\sigma_{yy})/2$,
radius $= \sqrt{((\sigma_{xx}-\sigma_{yy})/2)^2 + \sigma_{xy}^2}$,
$\sigma_{1,2} = $ centre $\pm$ radius, $\theta_p =
\tfrac{1}{2}\arctan2(2\sigma_{xy}, \sigma_{xx}-\sigma_{yy})$,
$\tau_{\max} = $ radius. Use \texttt{atan2}, not \texttt{atan}, to put
the principal angle in the correct quadrant. Verification states:
(i) $(80,40,30)\to(96.1, 23.9, 28.2^{\circ}, 36.1)$; (ii) uniaxial
$(100,0,0)\to(100, 0, 0^{\circ}, 50)$; (iii) pure shear $(0,0,50)\to
(50, -50, 45^{\circ}, 50)$. A correct program reproduces all three.
\end{exercise}

\begin{exercise}
Implement a yield-criterion comparison: given a stress state
$(\sigma_{xx}, \sigma_{yy}, \sigma_{zz}, \sigma_{xy}, \sigma_{yz},
\sigma_{zx})$ and a yield stress $\sigma_y$, return whether the
state has yielded by Tresca and whether it has yielded by von Mises.
Test on five stress states chosen to highlight the differences
between the criteria.
\paragraph{Solution sketch.} The reference is
\texttt{code/yield\_criteria.py}: assemble the symmetric $3\times 3$
tensor, take eigenvalues (\texttt{numpy.linalg.eigvalsh}) for the
principal stresses, compute $\sigma_{\text{Tr}} = \sigma_1 - \sigma_3$
and $\sigma_{\text{vM}} = \sqrt{\tfrac{1}{2}\sum(\sigma_i-\sigma_j)^2}$,
and compare each to $\sigma_y$. Discriminating test states: uniaxial
(criteria agree), pure shear (largest disagreement, ratio $2/\sqrt3$),
equibiaxial (agree), biaxial tension-compression (large disagreement),
and pure hydrostatic (neither yields, since both equivalent stresses
are zero). The hydrostatic case is the instructive one: an arbitrarily
large equal-on-all-sides stress causes no yield because there is no
shear and no deviatoric component.
\end{exercise}

\begin{exercise}
Simulate the elastic stress field around a circular hole in a plate
under uniaxial tension using a finite-difference or finite-element
method. Compare the maximum stress at the hole's edge to the Kirsch
prediction $K_t = 3$. Vary the plate-to-hole size ratio and identify
the smallest ratio at which the infinite-plate approximation is
valid to within $10\,\%$.
\paragraph{Solution sketch.} Mesh a quarter-plate (exploiting the two
symmetry planes) with a refined ring of elements at the hole edge;
apply uniform far-field tension on the outer edge and symmetry
boundary conditions on the cut planes. Read $\sigma_{yy}$ at the edge
point perpendicular to the load and divide by the applied stress to get
the numerical $K_t$. For a wide plate the result converges to $3.0$;
the finite-width correction (Howland) raises $K_t$ as the hole grows
relative to the plate width $w$. Yield: for $d/w \lesssim 0.2$ (plate
at least five hole-diameters wide) the infinite-plate value is within
about $10\,\%$; below that ratio the finite-width amplification matters.
Mesh convergence must be checked by halving the edge element size until
$K_t$ stops moving.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a steel hanger rod to suspend a $1{,}000\,\si{\kilogram}$ load
from a structural beam. Specify the rod diameter using a factor of
safety of $2$ against yield, with $\sigma_y = 350\,\text{MPa}$. State
the stress in the rod at the design load.
\paragraph{Solution sketch.} Load $W = 1000(9.81) = 9810\,\text{N}$.
Allowable stress $\sigma_{\text{allow}} = \sigma_y/\text{FoS} = 350/2 =
175\,\text{MPa}$. Required area $A = W/\sigma_{\text{allow}} = 9810/
175\times 10^6 = 5.6\times 10^{-5}\,\si{\meter}^2 = 56\,\si{\milli
\meter}^2$. Required diameter $d = \sqrt{4A/\pi} = \sqrt{4(56)/\pi} =
8.4\,\si{\milli\meter}$; round up to a standard $10\,\si{\milli\meter}$
rod (or $M12$ if threaded, sizing on the root area). At $d = 10\,\si{
\milli\meter}$, $A = 78.5\,\si{\milli\meter}^2$ and the actual stress is
$\sigma = 9810/78.5\times 10^{-6} = 125\,\text{MPa}$, a true factor of
safety of $350/125 = 2.8$. A complete answer notes the thread root if
the rod is threaded and the connection detail at both ends.
\end{exercise}

\begin{exercise}
Design a cylindrical thin-walled pressure vessel to contain gas at
$p = 1.5\,\text{MPa}$ with internal diameter $0.8\,\si{\meter}$ and a
factor of safety of $3$ against yield, with $\sigma_y = 240\,\text{
MPa}$. State the wall thickness and the dominant stress.
\paragraph{Solution sketch.} Dominant stress is the hoop stress
$\sigma_{\theta} = pr/t$ with $r = 0.4\,\si{\meter}$. Allowable stress
$\sigma_{\text{allow}} = 240/3 = 80\,\text{MPa}$. Set $pr/t =
\sigma_{\text{allow}}$: $t = pr/\sigma_{\text{allow}} = (1.5\times 10^6)
(0.4)/(80\times 10^6) = 7.5\times 10^{-3}\,\si{\meter} = 7.5\,\si{\milli
\meter}$. Specify $t = 8\,\si{\milli\meter}$ (next standard plate),
giving a working hoop stress of $pr/t = 75\,\text{MPa}$ and a true FoS
of $3.2$. The complete answer checks the thin-wall validity ($t/r =
0.02 \ll 1$, valid), adds corrosion allowance and weld efficiency
factors per the relevant code, and notes that the hoop stress, twice
the axial, governs and dictates a longitudinal-seam weld of full
quality.
\end{exercise}

\begin{exercise}
Redesign a $20\,\si{\milli\meter}$-diameter shaft that steps down to
$15\,\si{\milli\meter}$ diameter without a fillet (sharp shoulder),
under bending and torsion that produce a nominal stress of $80\,
\text{MPa}$ at the small section. Choose a fillet radius to keep the
peak stress below $\sigma_y = 250\,\text{MPa}$. State the $K_t$ for
the chosen radius and the resulting peak stress.
\paragraph{Solution sketch.} Nominal stress $80\,\text{MPa}$; the
sharp shoulder has a very high $K_t$ and an unbounded elastic peak. To
hold the peak below $250\,\text{MPa}$ the allowable $K_t$ is $250/80 =
3.1$, which is generous, but good practice targets $K_t < 2$. From a
fillet chart (or figure \ref{fig:vol03ch03:fillet-kt}) for $D/d =
20/15 = 1.33$, a fillet radius giving $r/d \approx 0.1$, i.e. $r =
1.5\,\si{\milli\meter}$, yields $K_t \approx 1.7$. Peak stress
$\sigma_{\max} = K_t\sigma_{\text{nom}} = 1.7(80) = 136\,\text{MPa}$,
well below yield and below the $250\,\text{MPa}$ requirement with a true
margin of $1.8$. The complete answer specifies the fillet on the
drawing, notes that the gain costs almost nothing in size, and warns
that a machinist cutting a sharp corner instead would silently double
the peak stress.
\end{exercise}

\begin{exercise}
Design a bolted lap joint to carry a tensile force of $60\,\text{kN}$
between two $10\,\si{\milli\meter}$-thick steel plates, using two
$M20$ high-strength bolts. State the bolt spacing, the edge distance,
and the plate width, with checks against bolt shear, bearing, net-
section tension, and block shear.
\paragraph{Solution sketch.} Per-bolt shear load $= 60/2 = 30\,\text{
kN}$. (1) Bolt shear: an $M20$ grade 8.8 bolt has a shear area $\approx
245\,\si{\milli\meter}^2$ and a shear strength near $375\,\text{MPa}$,
capacity $\approx 92\,\text{kN}$ per bolt, ample. (2) Bearing on the
$10\,\si{\milli\meter}$ plate: bearing area $= d t = 20(10) = 200\,\si{
\milli\meter}^2$; at a bearing allowable of, say, $400\,\text{MPa}$,
capacity $80\,\text{kN}$, ample. (3) Net-section tension: with two
$22\,\si{\milli\meter}$ holes the net width must carry $60\,\text{kN}$
at the steel allowable; a plate width near $80\,\si{\milli\meter}$ keeps
the net-section stress under $\sigma_{\text{allow}}$. (4) Block shear:
set edge distance $\approx 1.5 d = 30\,\si{\milli\meter}$ and bolt
pitch $\approx 3 d = 60\,\si{\milli\meter}$ so the tear-out block has
enough shear-plus-tension area. Standard detailing rules (edge $\geq
1.5d$, pitch $\geq 2.5$--$3d$) satisfy all four checks here; the answer
states the governing check (net-section tension at this width) and the
resulting plate width.
\end{exercise}

\begin{exercise}
Design a thread for a fastener carrying $100\,\text{kN}$ axial tensile
load, with a factor of safety of $3$ against yield at the thread
root. Specify the nominal diameter, the thread pitch, and the
material grade. State the $K_t$ assumed at the thread root.
\paragraph{Solution sketch.} Choose grade 10.9 alloy steel, $\sigma_y
\approx 900\,\text{MPa}$, allowable $= 900/3 = 300\,\text{MPa}$.
Required root area for the static check $A_r = N/\sigma_{\text{allow}} =
100{,}000/300\times 10^6 = 3.3\times 10^{-4}\,\si{\meter}^2 = 333\,\si{
\milli\meter}^2$, root diameter $\sqrt{4(333)/\pi} \approx 21\,\si{
\milli\meter}$, so an $M24$ coarse thread (pitch $3\,\si{\milli\meter}$,
tensile-stress area $\approx 353\,\si{\milli\meter}^2$) meets the static
requirement. For static ductile loading the thread-root $K_t$
(typically $K_t \approx 3$) is largely relieved by local yielding and
need not be applied to the nominal-stress static check, but for fatigue
it governs and the answer must state that a fatigue check at $K_f
\approx 3$ would require a larger size or pre-tension. Specify nominal
$M24$, coarse pitch $3\,\si{\milli\meter}$, grade 10.9, and note the
static-versus-fatigue distinction explicitly.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A steel bracket has fractured cleanly through its narrowest section
at a sharp $90^{\circ}$ internal corner. Identify the failure mode
and propose the design change that would have prevented it.
\paragraph{Solution sketch.} A clean fracture through the narrowest
section at a sharp internal corner points to stress concentration:
the corner amplifies the nominal section stress (a sharp re-entrant
corner has a very high or theoretically unbounded elastic $K_t$), and
the part broke there because the local stress reached the material's
strength while the average section stress was still modest. If the
fracture surface is flat and crystalline the mode is brittle fracture
or final fast fracture; beach marks would indicate fatigue initiating
at the corner. The design fix is to add a generous fillet radius at the
corner to bring $K_t$ below $2$, and, if loading is cyclic, to combine
the fillet with a smoother section transition. The lesson is the
chapter's refrain: sharp internal corners are the most expensive
feature in mechanical design.
\end{exercise}

\begin{exercise}
A bolted joint shows visible elongation of the bolt holes in the
direction of the applied load, without bolt-shear failure. Identify
the failure mode and state which dimension of the joint was
under-designed.
\paragraph{Solution sketch.} Ovalised holes in the load direction with
intact bolts is the signature of bearing failure: the bolt has crushed
and yielded the plate material ahead of it, $\sigma_{\text{bearing}} =
P/(dt)$ exceeding the plate's bearing capacity. The under-designed
dimension is the plate thickness $t$ (equivalently the bolt diameter
$d$ is too small for the plate), since bearing capacity scales with the
product $dt$. The fix is a thicker plate, a larger bolt, or more bolts
to share the load. Bearing is a ductile, visible, progressive mode, in
contrast to the sudden net-section or block-shear fractures, which is
why the joint deformed rather than parted.
\end{exercise}

\begin{exercise}
A welded I-beam connection has cracked along the weld toe at the
flange-to-web junction. State whether the failure mode is brittle
fracture, fatigue, or yield, given that the structure was in
intermittent service for several years.
\paragraph{Solution sketch.} Several years of intermittent service plus
a crack at a weld toe (a classic stress-concentration and
residual-stress site) points to fatigue. Yield would have produced
gross plastic distortion, not a localised crack; a one-time brittle
fracture would more likely have occurred early, at an overload event,
rather than after years of cycling. The diagnostic confirmation is the
fracture surface: fatigue shows beach marks (concentric arrest lines)
radiating from the weld toe, with a final fast-fracture zone. The weld
toe combines a geometric notch, tensile residual stress, and possible
undercut defects, making it the canonical fatigue initiation site in
welded structures.
\end{exercise}

\begin{exercise}
A cantilever beam loaded at its free end has visibly bent at a
location near (but not at) the clamped end. Identify which loading
condition (tip load, intermediate load, distributed load, sustained
load) is most consistent with the observed deformation pattern.
\paragraph{Solution sketch.} For a tip load on a cantilever the bending
moment is largest at the clamp and falls linearly to zero at the tip,
so first yield (and visible permanent bend) should appear at the clamp
itself, not just inside it. A visible bend near but not at the clamp
indicates the maximum moment, hence the peak stress, occurred slightly
inboard of the support. That happens when the load is applied at an
intermediate point rather than the free tip (the moment peaks at the
load point and the support carries the largest moment only if the load
reaches the tip), or when there is a local section reduction or defect
just inside the clamp. The pattern is most consistent with an
intermediate load combined with a section or material weak spot at the
yielded location; a pure tip load would yield exactly at the clamp.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Re-read the NBS report on the Hyatt Regency walkway collapse
\cite{acc:nbs-hyatt-1982}. State the elastic stress concentration at
the bolt-to-box-beam interface of the modified upper connection,
the gross-section stress in the connection, and the resulting peak
local stress. Compare to the material's yield stress at the
location.
\paragraph{Solution sketch (rubric).} A complete answer reports, from
the NBS report \cite{acc:nbs-hyatt-1982}: that the original single-rod
design was changed to a two-rod hang, doubling the load on the upper
walkway's box-beam-to-rod connection; that the load was transferred
through the threaded rod bearing on the box-beam weld and washer rather
than through the beam's full section; and that the connection's capacity
was roughly half the load it actually carried at collapse. The
mechanism is a stress concentration at the rod-nut-to-box-beam
interface combined with the doubled load: the welded box-beam seam and
the bearing of the nut concentrate stress where the section is weakest.
The model answer states the gross-section stress on the box beam, the
concentration at the weld and bearing point, and the conclusion that
the peak local stress exceeded the connection's capacity by roughly a
factor of two at the design load, before any live-load surge. The grade
hinges on tracing the failure to the connection detail and the load
doubling, not to a global overload.
\end{exercise}

\begin{exercise}
A pressure vessel of $4\,\si{\meter}$ diameter and $20\,\si{\milli
\meter}$ wall fails by tearing along a longitudinal weld during a
hydrostatic test at $1.2\,\text{MPa}$. State, in $200$ to $300$
words, the most probable failure mode (hoop-stress overload at a
weld defect, fatigue, brittle fracture in the weld heat-affected
zone, or other), and what evidence at the failure surface would
distinguish between them.
\paragraph{Solution sketch (rubric).} First the numbers: hoop stress
$\sigma_{\theta} = pr/t = (1.2\times 10^6)(2.0)/(0.020) = 120\,\text{
MPa}$, a modest stress for steel, so a clean hoop overload of sound
material is unlikely. A first-application hydrostatic test that tears
along a longitudinal weld most probably reveals a pre-existing weld
defect (lack of fusion, slag, undercut, or a hydrogen crack in the
heat-affected zone) at which the local stress was concentrated above
the nominal $120\,\text{MPa}$. Distinguishing evidence: a brittle
HAZ fracture shows a flat, crystalline, often chevron-marked surface
pointing back to the origin defect, with little wall thinning; a
ductile hoop overload shows shear lips and visible wall thinning at the
tear; fatigue (unlikely on a first test) would show beach marks. The
strong answer computes the hoop stress, notes it is low, and therefore
locates the cause at a weld flaw, naming the surface features that would
confirm it.
\end{exercise}

\begin{exercise}
A steel rotating shaft of $50\,\si{\milli\meter}$ diameter has failed
in service after several million cycles, with the fracture initiating
at a $5\,\si{\milli\meter}$-deep keyway. State the elastic $K_t$ at
the keyway corner, the dominant failure mode (fatigue or static
overload), and the calculation that a design review should have
performed to catch the inadequate design.
\paragraph{Solution sketch.} A keyway with a sharp corner in a shaft
under bending or torsion has an elastic $K_t$ in the range of about
$2$ to $4$ depending on the corner radius (a standard square keyway in
torsion is often taken as $K_t \approx 3$ to $3.5$). ``Several million
cycles'' is the unambiguous signature of fatigue, not static overload:
static failure happens once, on the first overload, not after millions
of cycles. The design-review calculation that would have caught it: take
the nominal alternating stress on the shaft from the rotating bending or
torque, multiply by the fatigue concentration $K_f$ at the keyway, and
compare the result against the material's endurance limit (corrected for
surface finish, size, and reliability) with a fatigue factor of safety.
The omission is almost always the failure to apply $K_f$ at the keyway,
which turns a nominal stress comfortably below the endurance limit into
an effective stress above it. The fix is a rounded keyway corner, a
larger shaft, or a different torque-transmission feature (spline,
shrink fit) with a lower concentration.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer argues that, since the part is ductile, the
stress concentration at the fillet does not matter and gross-section
stress is sufficient for design. Write a one-paragraph reply that
states when the argument is acceptable and when it fails.
\paragraph{Solution sketch (model answer).} The argument is partly
right and dangerously incomplete. For a ductile metal under static,
monotonic loading, local yielding at the fillet redistributes stress
and the concentration is largely relieved, so designing on gross-section
stress with a sensible factor of safety is defensible. The argument
fails for three loading types the junior engineer omitted: fatigue
(cyclic loading initiates cracks at the concentration even when the
nominal stress is low, and ductility does not save the part), brittle
or low-temperature service (no plastic redistribution is available), and
impact or high strain rate (the material behaves more brittlely). The
correct reply is: yes for ductile static loading, no for fatigue,
brittle materials, low temperature, or impact, and when in doubt apply
the concentration.
\end{exercise}

\begin{exercise}
A senior colleague proposes to size a critical component using
average stress on the cross-section, on the grounds that the local
stress field is too uncertain to compute. Write a one-paragraph
reply that takes the argument seriously and identifies the
specific design situations in which gross-section stress is
acceptable.
\paragraph{Solution sketch (model answer).} The colleague's concern is
legitimate: detailed local stress fields at a notch carry modelling
uncertainty, and over-precise computation can give false confidence.
Average gross-section stress is acceptable when the component is
ductile, statically loaded, free of crack-like defects, and detailed
with generous fillets so any concentration is mild, with a conservative
factor of safety covering the residual uncertainty. It is not
acceptable for fatigue, brittle materials, sharp geometric features, or
safety-critical parts, where the concentration governs and must be
bounded by tabulated $K_t$ values or finite-element analysis. The
defensible position handles the uncertainty with a documented factor of
safety and conservative detailing rather than by ignoring the
concentration; a strong answer reconciles the two views rather than
overruling the colleague outright.
\end{exercise}

\begin{exercise}
A designer asks whether Tresca or von Mises is the right yield
criterion for a new aluminum airframe component. Write a one-
paragraph reply that states the considerations and gives a defensible
recommendation.
\paragraph{Solution sketch (model answer).} For a ductile aluminium
alloy under combined loading where weight is at a premium, von Mises is
the defensible default: it matches experimental yield of ductile metals
more closely than Tresca and avoids the up-to-fifteen-percent
conservatism that would add unnecessary mass to an airframe. Von Mises
is also the criterion built into virtually all structural finite-element
post-processing, so it integrates with the analysis tools the team will
use. The considerations that could push toward Tresca are a governing
design code that mandates it, a desire for extra conservatism on a
safety-critical primary structure, or a hand-calculation context where
Tresca's $\sigma_1 - \sigma_3$ is simpler. The recommendation: use von
Mises for the airframe component, while confirming that the airworthiness
code in force permits it and applying the code's required factors of
safety on top.
\end{exercise}

\begin{exercise}
Identify a manufactured part in your immediate environment whose
critical stress concentration you have never identified. State, in
$200$ to $300$ words, the chosen part, the loading at the
concentration, the estimated $K_t$, and the failure mode the
concentration would govern.
\paragraph{Solution sketch (rubric).} There is no single answer; a
strong response picks a real part and reasons concretely. A good
example: the snap hook on a key ring, loaded by the weight and snag
forces on the keys, with a sharp internal corner where the gate meets
the body giving $K_t \approx 3$; the governing mode is fatigue at that
corner after repeated opening, eventually cracking the gate. Another:
the eyeglass-arm hinge, a thin steel pin in bending, $K_t \approx 2$ at
the pin shoulder, failing by fatigue after thousands of fold cycles.
Grade on whether the writer (1) names a specific part and its real
loading, (2) locates the actual concentration feature (hole, fillet,
thread, notch, corner), (3) estimates a plausible $K_t$ from the chapter's
canonical values, and (4) identifies the failure mode the concentration
would drive (fatigue for cyclic parts, brittle fracture for brittle
materials). The point of the exercise is to make the reader see
concentrations everywhere, which is the working engineer's habit of
mind.
\end{exercise}
