\chapter{Metabolism and bioenergetics}
\label{vol06:ch02}

\epigraph{The energy required by the higher animals is supplied by the
oxidation of the food, with the formation of carbonic acid and water,
and the liberation of heat.}{Hermann von Helmholtz, \emph{On the
Conservation of Force} (1862--1863), in the English translation
\cite[p.~210]{hist:helmholtz1863}.}

\chapmeta{Half-life: foundational. Archetypes: balance, transport. Exercise target: 30. Project track: estimation exercise (three independent estimates of an adult human's daily ATP turnover, reconciled to within a factor of two). Hazard class: none. Reader path default: standard, with sections explicitly tagged. Named cases: none required; metabolic-disease examples are described as mechanism classes.}

\noindent
A cell maintains its disequilibrium by continuous expenditure of free
energy. The chemical currency of that expenditure is adenosine
triphosphate (ATP), and every other claim Volume VI will make about
biological energy traces back to the ATP economy. The chapter
quantifies the cell's metabolic accounting: it identifies ATP as the
universal currency, walks through the central catabolic and anabolic
pathways, treats photosynthesis as the source that funds the rest of
the food web, develops metabolic flux analysis as an accounting tool,
and closes by naming three failure modes in which the metabolic
balance comes apart. Lavoisier's 1789 calorimeter on a guinea pig
established the quantitative identity of respiration and slow
combustion \cite{hist:v6c02-lavoisier-respiration-1789}; the
intervening two centuries have, at the level of accounting,
substantiated the identification and refined the bookkeeping. The
present chapter inherits Lavoisier's accounting stance.

\section{ATP as the universal currency}\pathtag{core}

\engterm{Adenosine triphosphate}. ATP is a nucleotide consisting of
the base adenine, the sugar ribose, and a chain of three phosphate
groups linked by anhydride bonds. The hydrolysis of the terminal
phosphate bond,
\[
\text{ATP} + \text{H}_2\text{O} \longrightarrow \text{ADP} + \text{P}_i,
\]
releases free energy. The standard Gibbs free energy of hydrolysis at
$\text{pH}\,7.0$, $310\,\si{\kelvin}$, and unit activity of all
species is $\Delta G^{\circ\prime} = -30.5\,\text{kJ}/\text{mol} =
-7.3\,\text{kcal}/\text{mol}$ \cite[ch.~13]{text:lehninger-biochemistry}.
Under physiological conditions, where the ratio
$[\text{ATP}]/[\text{ADP}][\text{P}_i]$ is maintained far from
equilibrium, the actual free energy of hydrolysis is approximately
$-50$ to $-60\,\text{kJ}/\text{mol}$, almost twice the standard value.
The corrected value is the engineering value: it is the work that the
cell can extract per ATP hydrolysed. Figure~\ref{fig:vol06:ch02:atp-bar}
plots ATP against its higher- and lower-energy neighbours on a common
free-energy axis; the intermediate position is the structural reason
ATP can act as a currency rather than as either a source or a sink.

\input{volumes/06-life/ch02-metabolism-and-bioenergetics/figures/fig-atp-hydrolysis-bar}

The cell's ATP pool turns over rapidly. A typical animal cell
maintains $1$ to $10\,\text{mM}$ ATP and turns over its entire pool
in seconds. An adult human, at rest, hydrolyses approximately
$3 \times 10^{20}$ ATP molecules per second, which corresponds to
a turnover of approximately $50\,\text{kg}$ of ATP per day, more
than half the body's mass. The turnover number is not a contradiction:
the cell does not have to synthesise $50\,\text{kg}$ of ATP from
scratch each day, because each ATP is hydrolysed to ADP, the ADP is
re-phosphorylated to ATP, and the cycle repeats. The body's
\emph{net} ATP synthesis from scratch is small (a few grams per day
to maintain the steady-state pool); the \emph{gross} cycling rate is
the $50\,\text{kg}$ that the project at the end of the chapter will
estimate three ways.

\engterm{Why ATP}. The cell's choice of ATP as a universal energy
carrier is not arbitrary. The hydrolysis free energy ($-30\,\text{kJ}/
\text{mol}$ standard, $-50\,\text{kJ}/\text{mol}$ physiological) sits
in the middle of the range of biochemically useful energies. The
high-energy phosphate compounds in the cell span approximately
$15$ to $60\,\text{kJ}/\text{mol}$ of hydrolysis energy; ATP at
$30\,\text{kJ}/\text{mol}$ can phosphorylate the lower-energy
compounds (for example, glycerol, $-9.2\,\text{kJ}/\text{mol}$;
glucose-6-phosphate, $-13.8\,\text{kJ}/\text{mol}$) and be regenerated
from the higher-energy compounds (for example, phosphoenolpyruvate at
$-61.9\,\text{kJ}/\text{mol}$, creatine phosphate at
$-43.1\,\text{kJ}/\text{mol}$). The intermediate position lets ATP
act as a buffer between the energy sources and the energy sinks
\cite[ch.~13]{text:lehninger-biochemistry}. The complete table of
standard values appears in \texttt{data/standard\_dg\_cellular\_reactions.csv}
and is the working reference for the calculation exercises.

\engterm{Other energy carriers}. ATP is the dominant currency but not
the only one. \engterm{GTP} is structurally and energetically similar
to ATP and is used in protein synthesis and in a few signalling
pathways. \engterm{NADH and NADPH} are reduced cofactors that carry
electrons (and the free energy of those electrons) between reactions;
they are the input fuel for oxidative phosphorylation. \engterm{Acetyl-CoA}
carries a two-carbon acyl group plus the free energy of the thioester
bond. \engterm{FADH$_2$} is a reduced flavin cofactor with a similar
role to NADH at a slightly less negative potential. The reader can
think of the cell as running on multiple denominations of energy
currency, with conversion rates set by enzyme-catalysed exchange
reactions. The total free-energy flow through the cell is the sum
across all carriers; ATP is the carrier through which the largest
fraction passes. A useful conversion mnemonic: $1$ NADH carries
approximately $220\,\text{kJ}/\text{mol}$ of free energy down to the
$\text{O}_2/\text{H}_2\text{O}$ couple, which the cell partitions into
approximately $2.5$ ATP at the synthase, so each NADH is worth $\sim
130\,\text{kJ}/\text{mol}$ in cellular ATP units and $\sim 90\,\text{kJ}/
\text{mol}$ is dissipated as heat along the chain.

\engterm{The energy charge}. The relative concentrations of ATP, ADP,
and AMP characterise the cell's energy state by the energy charge,
defined as
\[
\text{EC} = \frac{[\text{ATP}] + \frac{1}{2}[\text{ADP}]}{[\text{ATP}] + [\text{ADP}] + [\text{AMP}]}.
\]
A cell in a fully phosphorylated state has $\text{EC} = 1$; a cell
in a fully dephosphorylated state has $\text{EC} = 0$. Healthy cells
hold $\text{EC} \approx 0.85$ to $0.95$. The energy charge regulates
the activity of many central metabolic enzymes: high charge
allosterically inhibits glycolysis and TCA-cycle entry (so the cell
does not waste catabolic capacity when it has plenty of ATP); low
charge stimulates them. The energy charge is one of the cell's
primary state variables and a useful summary of its metabolic
condition.

\engterm{Concentration ratios and the mass-action correction}. The
gap between $\Delta G^{\circ\prime} = -30.5$ and the cellular $\Delta G
\approx -55\,\text{kJ}/\text{mol}$ is mass-action: the cell holds
$[\text{ATP}]/[\text{ADP}][\text{P}_i]$ far from $1\,\text{M}^{-1}$.
For typical human muscle cytosol at rest, $[\text{ATP}] \approx
8\,\text{mM}$, $[\text{ADP}] \approx 9\,\mu\text{M}$ (the resting
free ADP is much lower than the total because most ADP is bound to
creatine kinase), and $[\text{P}_i] \approx 4\,\text{mM}$. The
mass-action ratio is $Q = [\text{ADP}][\text{P}_i]/[\text{ATP}] \approx
4.5 \times 10^{-6}\,\text{M}$, and the correction is $RT \ln Q \approx
-31\,\text{kJ}/\text{mol}$, yielding $\Delta G \approx -61.5\,
\text{kJ}/\text{mol}$. The same logic explains why the body's
phosphagen system (creatine phosphate) provides only the first few
seconds of maximum-intensity work: the creatine-phosphate hydrolysis
is rapid and exergonic, but the pool is small ($\sim 25\,\text{mM}$
in human muscle) and is exhausted before slower catabolic pathways
can take over.

\engterm{Why phosphate, and why three of them}. The triphosphate
chain (rather than a mono- or diphosphate) is a structural choice:
the second and third phosphoanhydride bonds are similar in energy,
but they are kinetically protected from hydrolysis. ATP is
thermodynamically high-energy and kinetically slow; uncatalysed
hydrolysis of ATP in pure water has a half-life of years. The cell
therefore deposits a large free-energy reserve in ATP that is
released only when a specific enzyme couples ATP hydrolysis to a
target reaction. The combination of thermodynamic favourability and
kinetic stability is the same engineering principle behind a
charged battery: large potential energy held by a high activation
barrier, with the discharge gated by a controlled circuit. The cell's
ATPases are the controlled circuit, and most of them couple hydrolysis
to mechanical, transport, or biosynthetic work with high efficiency.

\section{Glycolysis, TCA, and oxidative phosphorylation}\pathtag{core}

The principal catabolic pathway by which animal cells extract free
energy from glucose has three stages: glycolysis, the
tricarboxylic-acid (TCA) cycle, and oxidative phosphorylation. Each
stage produces ATP or reducing equivalents that the next stage
consumes. The full pathway extracts approximately $30$ to $32$ ATP
per glucose molecule. The per-substrate yields for glucose, palmitate,
ethanol, and other common fuels are tabulated in
\texttt{data/atp\_yield\_per\_substrate.csv} and plotted in
Figure~\ref{fig:vol06:ch02:fuel-yield}.

\engterm{Glycolysis}. Glycolysis is a ten-step pathway in the cytosol
that converts one glucose molecule (6 carbons) to two pyruvate
molecules (3 carbons each) with net production of two ATP and two
NADH. The first half of the pathway invests two ATP to phosphorylate
glucose twice and to cleave the resulting fructose-1,6-bisphosphate
into two three-carbon sugar phosphates; the second half pays back
four ATP by substrate-level phosphorylation. The two NADH carry
electrons that will enter oxidative phosphorylation downstream.
Glycolysis is universal: it appears, with minor variations, in
every cell that uses glucose as a fuel, from \emph{E.~coli} to
mammalian liver. The pathway predates eukaryotes and is the cell's
most evolutionarily conserved metabolic backbone.
Figure~\ref{fig:vol06:ch02:glycolysis} lists each enzyme and the
corresponding ATP / NADH ledger entry.

\input{volumes/06-life/ch02-metabolism-and-bioenergetics/figures/fig-glycolysis-ledger}

The regulation of glycolysis is concentrated at three irreversible
steps catalysed by hexokinase, phosphofructokinase-1 (PFK-1), and
pyruvate kinase. PFK-1 is the principal control point: it is
allosterically inhibited by high ATP and by citrate (a signal that the
TCA cycle has plenty of input) and is activated by AMP and by
fructose-2,6-bisphosphate (a signal that ATP is low and that more
glycolytic flux is needed). The regulation is a classical feedback
loop, with the upstream enzyme PFK-1 controlled by the downstream
products of the integrated pathway. The cellular reading is that
glycolysis runs at a rate that the cell's ATP demand sets, not at a
rate that glucose availability sets; the same logic governs most
catabolic pathways.

\engterm{Step-by-step free-energy balance}. The standard $\Delta
G^{\circ\prime}$ of the ten glycolytic steps does not, by itself,
explain the pathway's directionality: several steps (aldolase, TPI,
GAPDH, PGM, enolase) have standard free energies near zero or
positive. The pathway runs forward because the three irreversible
steps (hexokinase $-16.7$, PFK-1 $-14.2$, pyruvate kinase $-31.4\,
\text{kJ}/\text{mol}$ cellular) pull the rest of the pathway forward
by mass action: their products are continuously removed and their
substrates continuously supplied, so the near-equilibrium steps
between them adopt the small forward bias that the boundary
concentrations dictate. The same architecture (a few strongly
exergonic gated steps spaced through a chain of near-equilibrium
steps) appears throughout central metabolism. The pattern is the
biochemical analogue of a fluid network with a few pump stations
along a near-frictionless pipe: the pumps set the direction; the
pipe sets the impedance.

\engterm{The fate of pyruvate}. The pyruvate exiting glycolysis has
three principal fates depending on the cell type and the oxygen
availability. Under aerobic conditions in most cells, pyruvate is
imported into the mitochondrial matrix and converted to acetyl-CoA
by the pyruvate dehydrogenase complex, with loss of CO$_2$ and
production of NADH. Under anaerobic conditions in muscle and in some
microorganisms, pyruvate is reduced to lactate by lactate
dehydrogenase, regenerating the NAD$^+$ that glycolysis needs. In
yeast and some other microorganisms, pyruvate is converted to
ethanol and CO$_2$ by a two-step pathway (alcoholic fermentation).
The three fates correspond to the three principal anaerobic and
aerobic energy-extraction strategies that life uses. The branch point
is regulated by the redox state ($[\text{NADH}]/[\text{NAD}^+]$): when
the ratio is high, lactate dehydrogenase becomes favourable; when low,
pyruvate dehydrogenase wins.

\engterm{The TCA cycle}. The tricarboxylic-acid cycle (also Krebs
cycle, also citric-acid cycle) is an eight-step cyclic pathway in the
mitochondrial matrix that oxidises acetyl-CoA to two molecules of
CO$_2$ per cycle, with production of one GTP (equivalent to one ATP)
by substrate-level phosphorylation, three NADH, and one FADH$_2$. The
cycle is named after the three-carboxylate intermediate citrate, the
first product of acetyl-CoA condensation with oxaloacetate. The cycle
is amphibolic (it serves both catabolic and anabolic purposes):
intermediates are withdrawn for biosynthesis (succinyl-CoA for
porphyrin synthesis, $\alpha$-ketoglutarate for glutamate, oxaloacetate
for aspartate and glucose) and replenished by anaplerotic reactions
(pyruvate carboxylase converts pyruvate to oxaloacetate). The
regulation of the cycle is exerted at three steps (citrate synthase,
isocitrate dehydrogenase, $\alpha$-ketoglutarate dehydrogenase) by
the ratios $[\text{ATP}]/[\text{ADP}]$ and $[\text{NADH}]/[\text{NAD}^+]$
and by calcium availability. Figure~\ref{fig:vol06:ch02:tca} lays the
cycle out and the running ledger.

\input{volumes/06-life/ch02-metabolism-and-bioenergetics/figures/fig-tca-cycle}

The cycle produces one ATP, three NADH, and one FADH$_2$ per
acetyl-CoA, hence two ATP, six NADH, and two FADH$_2$ per original
glucose. Combined with the glycolytic and pyruvate-dehydrogenase
outputs, one glucose generates ten NADH, two FADH$_2$, four ATP
(two from glycolysis, two from the TCA cycle as GTP), and six CO$_2$
by the end of the catabolic phase. The NADH and FADH$_2$ are the
fuel for the third stage.

\engterm{Oxidative phosphorylation}. The mitochondrial inner membrane
houses the electron transport chain (ETC), a series of four
membrane-embedded protein complexes (Complex I, II, III, IV) that
transfer electrons from NADH and FADH$_2$ to molecular oxygen, with
formation of water. The transfer proceeds through a series of
oxidation-reduction reactions in which each complex captures the
free-energy drop across a portion of the chain and uses it to pump
protons from the matrix into the intermembrane space. The proton
gradient stores the captured free energy as a transmembrane
electrochemical potential difference $\Delta \mu_{\text{H}^+}$. The
fifth complex, ATP synthase, allows protons to flow back down the
gradient and uses the flow to drive the synthesis of ATP from ADP
and inorganic phosphate. Figure~\ref{fig:vol06:ch02:etc} shows the
membrane layout with proton-pumping stoichiometries.

\input{volumes/06-life/ch02-metabolism-and-bioenergetics/figures/fig-etc-chemiosmosis}

The chemiosmotic coupling hypothesis, proposed by Peter Mitchell in
1961, is the engineering reading of oxidative phosphorylation: the
membrane potential is the intermediate that couples electron transport
to ATP synthesis, the membrane is the capacitor that stores the
energy, and the ATP synthase is a molecular generator that converts
the proton-motive force into chemical work
\cite{paper:mitchell1961-chemiosmotic}. Mitchell's hypothesis won
the 1978 Nobel Prize and remains the canonical model. The proton-motive
force has two components, the membrane potential ($\Delta \Psi
\approx -150\,\text{mV}$, matrix negative) and the pH gradient
($\Delta \text{pH} \approx 0.5$ to $1$ unit, matrix alkaline), with
total chemical potential
\[
\Delta \mu_{\text{H}^+} = F \Delta \Psi - 2.303 RT \Delta \text{pH}
\approx -200\,\text{mV}\text{-equivalent}.
\]
ATP synthase uses approximately $3$ to $4$ protons per ATP
synthesised; the resulting ATP yield per NADH is approximately $2.5$
and per FADH$_2$ approximately $1.5$ (FADH$_2$ enters the chain at
Complex II, which does not pump protons, so it captures less of the
free-energy drop). Total ATP per glucose by oxidative phosphorylation
is approximately $10 \times 2.5 + 2 \times 1.5 = 28$, plus the four
ATP from substrate-level phosphorylation, for a total of $32$.

The actual yield depends on the cell type and the metabolic state.
The textbook value of $36$ to $38$ ATP per glucose (popular in older
biochemistry courses) assumes a higher P/O ratio than the modern
measurements support; the more accurate value is $30$ to $32$
\cite[ch.~19]{text:lehninger-biochemistry}. The c-ring stoichiometry
of the human mitochondrial ATP synthase has been resolved as eight
subunits per ring \cite{paper:v6c02-watt-stoichiometry-2010},
implying approximately $2.7$ protons per ATP synthesised (one rotation
of the rotor produces three ATP and translocates eight protons).
Complex I's stoichiometry is $4\,\text{H}^+$ per NADH oxidised
\cite{paper:v6c02-wikstrom-stoichiometry-2012}. Combined with $4$
H$^+$ at Complex III and $2$ H$^+$ at Complex IV, the chain pumps
$10$ H$^+$ per NADH, and the synthase consumes $\sim 2.7$ H$^+$ per
ATP, giving the modern $2.5$ ATP/NADH yield.

\input{volumes/06-life/ch02-metabolism-and-bioenergetics/figures/fig-fuel-yield}

\engterm{Other catabolic pathways}. Glucose is not the only fuel.
Fatty acids are catabolised by $\beta$-oxidation, a four-step cycle
in the mitochondrial matrix that removes two carbons at a time from
a fatty-acyl-CoA and feeds them into the TCA cycle as acetyl-CoA. A
$\text{C}_{16}$ palmitate yields $106$ ATP by complete oxidation,
compared to $32$ ATP for the same mass of glucose; fat is the
densest cellular fuel. Amino acids are catabolised through various
pathways that converge on acetyl-CoA, pyruvate, or TCA intermediates,
with the amino nitrogen disposed of as urea (in mammals) or ammonia
(in fish and many invertebrates). The volume's central engineering
reading is that catabolism is a network with multiple inputs and a
single common output (ATP), and that the cell switches input fuels
according to availability. The respiratory quotient
($\text{RQ} = \text{CO}_2$ produced $/ \text{O}_2$ consumed) of $1.00$
for pure carbohydrate, $0.70$ for pure fat, and $0.81$ for protein is
the standard indirect-calorimetry signal that a clinician reads to
infer substrate mix from breath gas analysis. The full table is at
\texttt{data/respiratory\_quotient\_by\_fuel.csv}.

\section{Photosynthesis}\pathtag{standard}

The free energy that funds the entire animal food web traces, at its
origin, to photosynthesis. Photosynthesis is the conversion of
electromagnetic energy (solar photons) into chemical energy (glucose
and other carbon-based fuels) by the photosynthetic organisms (plants,
algae, cyanobacteria). The overall reaction is
\[
6\,\text{CO}_2 + 6\,\text{H}_2\text{O} + \text{light} \longrightarrow
\text{C}_6\text{H}_{12}\text{O}_6 + 6\,\text{O}_2,
\]
with the light providing $\Delta G \approx +2870\,\text{kJ}/\text{mol}$
of glucose synthesised. The reaction is the engineering inverse of
the respiratory oxidation that the previous section described: where
respiration burns glucose to CO$_2$ and water with release of free
energy, photosynthesis uses light to drive the reverse synthesis with
capture of free energy.

\engterm{Light reactions}. The light-dependent reactions of
photosynthesis occur in the thylakoid membrane of the chloroplast.
The thylakoid contains two photosystems (PSII and PSI), each a
membrane-embedded protein complex with a chlorophyll-containing
reaction centre that absorbs a photon of approximately $680\,\text{nm}$
(PSII) or $700\,\text{nm}$ (PSI). The absorbed photon ejects an
electron from the chlorophyll, which then traverses an electron
transport chain analogous to the mitochondrial chain. Water is split
at PSII to replace the lost electrons, with release of O$_2$ (the
source of all atmospheric oxygen). The electron transport pumps
protons into the thylakoid lumen, generating a proton-motive force
that drives ATP synthesis through a chloroplast ATP synthase. The
final electron acceptor is NADP$^+$, reduced to NADPH at the end of
PSI. The Z-scheme of Figure~\ref{fig:vol06:ch02:z-scheme} plots the
reduction-potential trajectory and locates the two photochemical
uphill jumps.

\input{volumes/06-life/ch02-metabolism-and-bioenergetics/figures/fig-photo-z-scheme}

\engterm{Dark reactions (Calvin cycle)}. The Calvin cycle, in the
chloroplast stroma, uses the ATP and NADPH produced by the light
reactions to fix CO$_2$ into a three-carbon sugar (glyceraldehyde-3-phosphate)
that is then exported and converted to glucose, starch, or other
products. The cycle's rate-limiting enzyme is
ribulose-1,5-bisphosphate carboxylase/oxygenase (Rubisco), the most
abundant protein on Earth by mass: estimates of total biospheric
Rubisco mass are approximately $0.7\,\text{Gt}$, comparable to the
mass of all human-built infrastructure \cite[ch.~20]{text:lehninger-biochemistry}.
Rubisco is a notoriously slow enzyme (turnover number of $3$ to $10$
CO$_2$ molecules per second per active site, compared to $10^5$ or
more for fast enzymes); plants compensate by holding it at high
concentration ($0.5$ to $1.0\,\text{g}/\text{L}$ in chloroplast
stroma). The Calvin cycle's full stoichiometry, per six CO$_2$ fixed
to one glucose: $18$ ATP and $12$ NADPH consumed; the light reactions
must therefore deliver these in matched stoichiometry to keep the
ATP$/$NADPH ratio at $3:2$, an actively regulated balance.

\engterm{Efficiency}. The theoretical maximum efficiency of
photosynthesis, given the photon energies and the free-energy
requirement, is approximately $34\,\%$ at the photochemical level.
The actual whole-plant efficiency, integrated over the growing
season and accounting for non-photosynthetically-active radiation,
incomplete light absorption, photorespiration, dark respiration, and
non-leaf biomass costs, is $1\,\%$ to $4\,\%$ for the most
productive C4 crops (sugarcane, maize) and $0.5\,\%$ to $1\,\%$ for
typical C3 crops (wheat, rice, soybean) under field conditions
\cite[ch.~20]{text:lehninger-biochemistry,paper:v6c02-zhu-photosynthesis-2010}.
The actual efficiency is small, but plant photosynthesis is the only
large-scale solar-to-chemical conversion at industrial scale: the
global net primary production is approximately $100\,\text{Gt}$ of
carbon per year (current as of 2020), energy-equivalent to several
times humanity's annual primary energy consumption. The efficiency
cascade from quantum maximum to field-integrated value, with the
photochemical, biochemical, and ecological losses at each stage, is
summarised in \texttt{data/photosynthesis\_efficiency\_by\_crop.csv}
and is the topic of the chapter's second estimation block.

\engterm{C3, C4, CAM}. The Calvin cycle's first intermediate is the
three-carbon compound 3-phosphoglycerate, hence C3 photosynthesis.
The Rubisco enzyme that fixes CO$_2$ also accepts O$_2$ as a
substrate, producing 2-phosphoglycolate as a wasteful side product
(photorespiration); the side reaction is competitive at high
temperatures and high oxygen partial pressures. C4 plants
(approximately $3\,\%$ of plant species, including most tropical
grasses, maize, sugarcane, and many invasive weeds) have evolved an
additional carbon-fixation step that concentrates CO$_2$ around
Rubisco, suppressing photorespiration and increasing the
photosynthetic efficiency in hot, dry climates. CAM plants
(crassulacean acid metabolism, present in many succulents) open their
stomata at night, fix CO$_2$ into malate, store it until daylight,
and release it for the light-dependent reactions; CAM is an
adaptation to extreme water limitation. The three carbon-fixation
strategies form an evolutionary engineering family: each trades a
different cost (Rubisco selectivity, ATP cost of carbon pumping,
water loss) against a different gain.

\begin{estimation}
\textbf{Question.} Estimate the maximum theoretical efficiency of
photosynthesis from quantum considerations.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
\emph{Photon input.} A $680\,\text{nm}$ photon (red, at the PSII
absorption peak) carries energy $E_\gamma = hc/\lambda = (6.626 \times
10^{-34}) \cdot (3 \times 10^8) / (6.8 \times 10^{-7}) \approx 2.9
\times 10^{-19}\,\text{J}$, or $176\,\text{kJ}$ per mol of photons.

\emph{Photons per CO$_2$ fixed.} The Z-scheme requires $4$ electrons
per CO$_2$ (to produce one NADPH per electron pair, two NADPH per
CO$_2$). Each electron costs one photon at PSII and one at PSI:
$2$ photons per electron, $8$ photons per CO$_2$. Plus the ATP cost
of the Calvin cycle, which requires an additional $\sim 2$ photons
per CO$_2$ at the chloroplast ATP synthase. Total: $\sim 10$ photons
per CO$_2$, hence $60$ photons per glucose.

\emph{Quantum efficiency.} Glucose stores $2870\,\text{kJ}/\text{mol}$
of free energy. Input is $60 \times 176 = 10\,560\,\text{kJ}/\text{mol}$
of photons. Quantum efficiency is $2870 / 10\,560 \approx 27\,\%$.

\emph{Spectrum correction.} The full solar spectrum carries
approximately $45\,\%$ of its energy in the PAR ($400\,\text{nm}$ to
$700\,\text{nm}$) band; longer wavelengths cannot drive PSII or PSI.
The whole-spectrum-to-fixed-carbon maximum is $0.45 \times 0.27 \approx
12\,\%$.

\emph{Field correction.} Reflectance ($\sim 10\,\%$), non-leaf biomass
cost ($\sim 30\,\%$), photorespiration in C3 ($\sim 20\,\%$), and
dark respiration ($\sim 30\,\%$) reduce the value further to a field
maximum of $\sim 4\,\%$ for productive C4 crops and $\sim 1\,\%$ for
typical C3 field crops.

\smallskip
\textit{Postmortem.} The cascade explains why measured C4 field
efficiencies (sugarcane: $\sim 2.0\,\%$, maize: $\sim 1.5\,\%$) sit
where they do. The headroom between current and theoretical is
real (factor of $\sim 2$); crop-breeding and reactor-design programs
have targeted it for decades \cite{paper:v6c02-zhu-photosynthesis-2010}.
Closed algal photobioreactors have demonstrated $\sim 3\,\%$ to
$8\,\%$ under artificial illumination, but they pay for the gain in
electricity and capital cost. Outside the biological context, the
photovoltaic-then-electrolysis route now exceeds $12\,\%$
solar-to-fuel in the laboratory, several times the biological
maximum; photosynthesis is at last in serious competition with
inorganic alternatives.
\end{estimation}

\section{Metabolic flux analysis}\pathtag{standard}

The cell's metabolism is a network of hundreds of reactions, each
with its own rate. The engineer who wishes to predict cellular
behaviour, or to redesign it for biomanufacturing, needs an
accounting tool that handles networks of this size. The standard
tool is \engterm{metabolic flux analysis} (MFA) and its companion
\engterm{flux balance analysis} (FBA).

\engterm{Stoichiometric matrix}. Each reaction in the metabolic
network has a stoichiometry (one molecule of glucose consumed per
two molecules of pyruvate produced, etc.). The stoichiometries are
collected in a matrix $\mathbf{S}$ with $m$ rows (one per metabolite)
and $n$ columns (one per reaction). The entry $S_{ij}$ is the
stoichiometric coefficient of metabolite $i$ in reaction $j$ (positive
if produced, negative if consumed). At steady state the flux $v_j$
through each reaction satisfies the mass-balance equation
\[
\mathbf{S} \mathbf{v} = \mathbf{0}
\]
for all internal metabolites. The set of fluxes that satisfy the
balance constitutes the cell's feasible operating envelope. The
matrix $\mathbf{S}$ is typically large (the metabolic reconstruction
of \emph{E.~coli} contains $\sim 2500$ reactions and $\sim 1700$
metabolites) and sparse (each reaction involves only a few
metabolites). A six-reaction, four-metabolite toy example, with the
full $\mathbf{S}$ written out, appears in
Figure~\ref{fig:vol06:ch02:mfa-toy}.

\input{volumes/06-life/ch02-metabolism-and-bioenergetics/figures/fig-mfa-toy-network}

\engterm{Flux balance analysis}. FBA selects a particular flux vector
from the feasible envelope by maximising or minimising a chosen
objective function, typically biomass production for growing cells or
product yield for biomanufacturing. The optimisation is a linear
program because the constraints (mass balance plus upper and lower
bounds on individual fluxes) are linear. The classical formulation
is
\[
\max_{\mathbf{v}} c^T \mathbf{v} \quad \text{subject to} \quad
\mathbf{S} \mathbf{v} = \mathbf{0}, \quad v_j^{\min} \le v_j \le v_j^{\max}.
\]
The result is a single flux vector that the cell is predicted to
adopt under the chosen objective. The prediction is accurate to within
$10\,\%$ to $20\,\%$ for many central-carbon-metabolism fluxes in
\emph{E.~coli} under defined growth conditions
\cite[ch.~3]{text:alon-systems-biology,paper:v6c02-orth-cobra-2010}.
FBA is now the standard in-silico screening tool for metabolic
engineering in biomanufacturing. The script
\texttt{code/metabolic\_flux\_lp.py} solves the toy network of
Figure~\ref{fig:vol06:ch02:mfa-toy} analytically and (with scipy
available) verifies the answer using
\texttt{scipy.optimize.linprog}.

\engterm{Choice of objective}. The standard FBA objective is the
biomass-production flux, which models cell growth as a single
composite reaction whose stoichiometry encodes the mass fractions of
protein, DNA, RNA, lipid, and small molecules in the cell's dry
mass. The biomass coefficients are taken from experimentally
measured cellular composition (for \emph{E.~coli}, approximately
$55\,\%$ protein, $20\,\%$ RNA, $9\,\%$ lipid by dry mass). Different
objectives produce different predictions: minimising substrate
uptake at fixed biomass models a carbon-limited chemostat; maximising
ATP production models a stationary-phase cell; maximising a specific
product flux models a biomanufacturing strain. The choice of
objective is the modeller's responsibility, and the validity of the
prediction depends on the cell's actual operating regime matching
the assumed objective.

\engterm{$^{13}$C metabolic flux analysis}. The actual fluxes through
a metabolic network can be measured directly by feeding cells a
glucose substrate labelled with $^{13}$C at a known position and
tracing the labelling pattern through downstream intermediates. The
labelling distribution constrains the relative fluxes through
parallel pathways and resolves cycles that FBA cannot. $^{13}$C MFA
is the experimental complement to FBA and the basis of most
quantitative metabolic-engineering papers
\cite[ch.~3]{text:alon-systems-biology}. A typical $^{13}$C-MFA
experiment in \emph{E.~coli} uses a $1$-$^{13}$C glucose source,
samples the steady-state labelling of $\sim 30$ central metabolites
by mass spectrometry, and inverts the labelling pattern to a flux
distribution with uncertainty bounds. The data resolve, for example,
the partitioning between glycolysis and the pentose-phosphate pathway,
which FBA alone cannot determine because both pathways carry the same
input and produce the same output stoichiometry.

\engterm{Limitations}. MFA and FBA assume steady state and ignore
the dynamics of intermediate accumulation. They do not handle
regulatory dynamics, allosteric inhibition, or rapid transitions
(diauxic shifts, stationary-phase entry). Kinetic modelling of the
full network is technically possible but parameter-intensive; the
working compromise is to combine FBA for large-scale screening with
detailed kinetic models for the rate-limiting steps of the targeted
pathway. The standard handoff sequence in industrial metabolic
engineering is: FBA identifies the maximal theoretical yield;
$^{13}$C-MFA measures the current actual yield; the difference
indicates where the bottleneck lies; targeted kinetic modelling and
strain engineering attempt to close the gap.

\section{Free-energy balances in cells}\pathtag{standard}

The free-energy balance across a metabolic pathway is the engineering
fundamental of the previous sections, summarised in equation form.
For a reaction $A + B \rightleftharpoons C + D$ with free-energy
change $\Delta G^{\circ\prime}$ at standard conditions, the actual
free-energy change at physiological conditions is
\[
\Delta G = \Delta G^{\circ\prime} + RT \ln \frac{[C][D]}{[A][B]}.
\]
The reaction proceeds forward when $\Delta G < 0$ and reverses when
$\Delta G > 0$. The cell can drive a reaction with $\Delta G^{\circ\prime}
> 0$ forward at physiological concentrations by holding the reactant
concentrations high and the product concentrations low, or by coupling
the reaction to ATP hydrolysis. A common pattern is the
\engterm{coupling of an unfavourable reaction to ATP hydrolysis}: if
$A \rightarrow B$ has $\Delta G^{\circ\prime} = +20\,\text{kJ}/\text{mol}$,
the coupled reaction $A + \text{ATP} \rightarrow B + \text{ADP} +
\text{P}_i$ has $\Delta G^{\circ\prime} = -10\,\text{kJ}/\text{mol}$ and
proceeds forward. The cell pays one ATP per $A \rightarrow B$
conversion; the price is the engineering cost of running the cell's
chemistry uphill.

The pathway-level free-energy balance is the sum of the individual
reaction $\Delta G$ values. A pathway is energetically feasible at the
cellular concentrations if its overall $\Delta G < 0$. The pathway's
individual steps need not all be favourable: a few uphill steps can
be tolerated if they are flanked by downhill steps and if the overall
balance is negative. Glycolysis illustrates: most of its steps have
$\Delta G$ near zero at physiological concentrations, with the three
regulatory steps (hexokinase, PFK-1, pyruvate kinase) having
strongly negative $\Delta G$ that drives the overall flux forward.

\engterm{Coupling ratios}. The coupling ratios in the cell's energy
network are quantitative. Glycolysis pays $2$ ATP and produces
$4$ ATP plus $2$ NADH, a net gain of $2$ ATP plus the reducing
equivalents. Oxidative phosphorylation converts each NADH to $2.5$
ATP and each FADH$_2$ to $1.5$ ATP. The overall yield of glucose
oxidation is $30$ to $32$ ATP. The thermodynamic limit (the free
energy of complete glucose oxidation, $2870\,\text{kJ}/\text{mol}$,
divided by the physiological ATP synthesis energy,
$60\,\text{kJ}/\text{mol}$) is $48$ ATP per glucose. The actual yield
is $63\,\%$ to $67\,\%$ of the thermodynamic limit; the remaining
energy appears as metabolic heat (the body's primary heat source at
rest) and as the irreversibility of each catabolic step. The
efficiency is comparable to a well-designed industrial heat engine.

\engterm{Per-organ accounting}. The body's basal ATP demand is
unevenly distributed. Brain, liver, heart, and kidney consume more
energy per unit mass than any other tissue
\cite{paper:v6c02-elia-organ-bmr-1992}. The brain, at $\sim 2\,\%$
of body mass, accounts for $\sim 20\,\%$ of basal metabolic rate;
the kidneys, at $\sim 0.4\,\%$, account for $\sim 8\,\%$. Resting
skeletal muscle accounts for $\sim 20\,\%$ of BMR despite holding
$40\,\%$ of body mass, because the resting muscle's specific
metabolic rate is small. During exercise, the muscle fraction rises
sharply (a factor of $20$ to $30$ in working muscle) while the other
organs hold approximately constant; the body's total metabolic rate
can rise by a factor of $20$ above basal. The per-organ table is
recorded in \texttt{data/basal\_metabolic\_rate\_by\_organ.csv}.

\engterm{Where the ATP goes}. A second partition, by sink rather than
by source, is the breakdown of the cell's ATP consumption among its
principal user processes. Approximately $25\,\%$ of cellular ATP is
spent on protein synthesis (peptide-bond formation and tRNA charging),
$20\,\%$ on Na$^+$/K$^+$-ATPase ion pumping, $5\,\%$ on Ca$^{2+}$
pumping, $15\,\%$ on actin-myosin or microtubule mechanical work,
$10\,\%$ on gluconeogenesis and biosynthesis, and the rest on smaller
sinks (urea synthesis, lipid synthesis, mRNA degradation, signalling)
\cite{paper:v6c02-rolfe-brown-1997}. Approximately $20\,\%$ of the
mitochondrial proton flux is lost to basal proton leak (uncoupled
respiration that hydrolyses substrates without producing ATP)
\cite{paper:v6c02-jastroch-uncoupling-2010}; the leak contributes to
basal heat production and is a target for hypothesised
weight-management interventions. The cellular energy Sankey of
Figure~\ref{fig:vol06:ch02:sankey} sketches the partitioning.

\input{volumes/06-life/ch02-metabolism-and-bioenergetics/figures/fig-energy-sankey}

\engterm{Closing the budget against the second law}. The body's
sustained metabolic heat output ($\sim 100\,\text{W}$ for a sedentary
adult) is at steady state equal to the body's metabolic chemical
input rate; no net free energy accumulates internally. The detailed
budget reads: $\sim 100\,\text{W}$ of food in, $\sim 100\,\text{W}$ of
heat out, with the body's mass, composition, and temperature
unchanged. Any claim that violates this balance (a diet that produces
weight gain at zero caloric intake, a therapy that increases body
temperature without metabolic substrate, a supplement that
\engterm{boosts ATP} without measurable substrate consumption)
violates the first law and can be rejected without further analysis.
The failure section returns to the point.

\section{Worked examples}\pathtag{standard}

\engterm{Worked example 1: muscle contraction}. Skeletal muscle
generates mechanical work by hydrolysing ATP at actin-myosin
crossbridges. The molecular detail is in Chapter~\ref{vol06:ch04};
here we account for the energy budget. A maximally exercising human
hydrolyses ATP at $80$ to $100$ times the resting rate. The body's
ATP pool ($\sim 0.1\,\text{mol}$) would be exhausted in $1$ to
$2\,\text{seconds}$ if not regenerated. Three regeneration systems
operate at different timescales: creatine phosphate (capacity for a
few seconds of maximum work, rapid regeneration of ATP by transfer
of phosphate); anaerobic glycolysis (capacity for tens of seconds to
minutes of high-intensity work, with lactate accumulation); and
oxidative phosphorylation (sustainable indefinitely if fuel and
oxygen are available, the dominant pathway for endurance work). The
three systems' transitions appear in athletic performance as the
familiar split between sprint, anaerobic, and aerobic events. A
$100\,\text{m}$ sprint draws principally on creatine phosphate; a
$400\,\text{m}$ event on anaerobic glycolysis with substantial
lactate accumulation; a marathon entirely on oxidative phosphorylation
of fat (with a small carbohydrate margin to spare).

The mechanical efficiency of muscle, measured as the ratio of
mechanical work output to the chemical energy of the consumed
substrate, is $20\,\%$ to $25\,\%$ for steady aerobic work. The
remainder is dissipated as heat; the body's principal mechanism for
heat dissipation during exercise is evaporative cooling through
perspiration, with a peak rate of $1\,\text{L}/\text{h}$ for an
acclimatised endurance athlete. The body's heat-management system
fails (in heatstroke) when sweat-evaporative cooling cannot keep up
with metabolic heat production. The failure mode is one of the
chapter's named failure scenarios. A worked number: a $70\,\text{kg}$
runner at $300\,\text{W}$ mechanical output dissipates
$\sim 900\,\text{W}$ of metabolic heat (chemical input $\sim 1200\,
\text{W}$ at $25\,\%$ mechanical efficiency); sweat-evaporative cooling
at $2400\,\text{kJ}/\text{L}$ of evaporated water requires
$\sim 1.3\,\text{L}/\text{h}$ to keep pace, near the upper limit of
sustainable sweat rate.

\engterm{Worked example 2: fermentation}. Yeast fermentation
(\emph{Saccharomyces cerevisiae}) converts glucose to ethanol and
CO$_2$ in the absence of oxygen, with a net yield of $2$ ATP per
glucose. The industrial process produces ethanol at concentrations
up to $15\,\%$ to $18\,\%$ (above which the yeast itself is killed
by the product), with theoretical maximum yield of $0.511\,\text{g}$
ethanol per gram of glucose and typical industrial yields of
$0.45$ to $0.50\,\text{g}/\text{g}$ ($90\,\%$ to $98\,\%$ of theoretical)
\cite{paper:v6c02-thauer-anaerobe-1977}. The fermentation pathway is
the same glycolytic pathway that animal cells use, with two additional
steps (pyruvate decarboxylase and alcohol dehydrogenase) that
regenerate NAD$^+$ from NADH in the absence of oxygen. The economic
significance is large: bioethanol fuel and beverage production
together exceed $100\,\text{million tonnes}$ per year globally
(current as of 2024) \cite{text:lehninger-biochemistry}.
Chapter~\ref{vol06:ch10} on biomanufacturing will return to
fermentation as a process. The energy efficiency of the fermentation
route, measured per gram of glucose substrate, is much lower than
respiration ($2$ ATP per glucose versus $30$ to $32$); the
fermentation route survives because it produces a useful product
(ethanol) at a high mass yield rather than because it captures the
substrate's free energy.

\engterm{Worked example 3: photosynthesis efficiency}. A maize plant
under typical field conditions captures approximately $1$ to $2\,\%$
of incident photosynthetically active radiation in fixed-carbon
biomass over a growing season. With incident solar radiation of
$5\,\text{kWh}/(\text{m}^2 \cdot \text{day})$ at midlatitudes, of which
approximately $45\,\%$ is photosynthetically active, the absorbed
PAR is $2.25\,\text{kWh}/(\text{m}^2 \cdot \text{day})$. At $2\,\%$
efficiency, the chemical-energy capture is $0.045\,\text{kWh}/
(\text{m}^2 \cdot \text{day})$, or $160\,\text{kJ}/(\text{m}^2 \cdot
\text{day})$. Glucose contains $15.6\,\text{kJ}/\text{g}$ of energy,
so the daily biomass synthesis is $10\,\text{g}/(\text{m}^2 \cdot
\text{day})$, or $100\,\text{kg}/(\text{ha} \cdot \text{day})$. Over
a $120\,\text{day}$ growing season the cumulative biomass is
$12\,\text{t}/\text{ha}$ of glucose-equivalent. Maize field yields
of $10$ to $12\,\text{t}/\text{ha}$ of grain plus comparable stover
match the estimate. The same arithmetic for sugarcane (field
efficiency $\sim 2\,\%$, season $\sim 365$ days in tropical climates)
yields $\sim 80\,\text{t}/\text{ha} \cdot \text{year}$, matching the
documented record harvest.

\begin{archetype}[Balance and transport in the cell]
The cell exhibits the balance and transport archetypes in their most
quantitative form. The balance is the free-energy balance across every
catabolic and anabolic pathway: free energy is conserved across each
reaction (the second law of thermodynamics), and the cell's machinery
arranges the reactions to extract useful work at each step. The
transport is the carrier-coupled movement of substrates across
membrane barriers: every metabolite that enters or leaves an organelle
must cross at least one bilayer, and the cell pays an energy cost
(directly via ATP or indirectly via a coupled gradient) for the
selectivity that the membrane enforces. The two archetypes will
recur throughout Volume VI; the cell's metabolic network is the
textbook example.
\end{archetype}

\section{Failure: metabolic disease and thermodynamic hype}\pathtag{core}

Metabolic failure is the cell's failure to maintain its disequilibrium
against the demands of its environment. Four classes of failure
illustrate the volume's framing, three from medicine and one from
the recurring intersection of biology with engineering hype.

\engterm{Mitochondrial dysfunction}. The mitochondrion is the cell's
principal ATP producer; its failure cascades through every other
cellular function. Inherited mitochondrial diseases (Leber's hereditary
optic neuropathy or LHON, MELAS, Leigh syndrome) arise from mutations
in either the mitochondrial DNA (which encodes a subset of
respiratory-chain subunits and the mitochondrial tRNAs) or in
nuclear-encoded mitochondrial proteins \cite{paper:v6c02-wallace-1999}.
LHON is associated with point mutations in mitochondrial ND4 (the
m.11778G$>$A variant being the most common)
\cite{paper:v6c02-wallace-leber-1988}; MELAS most commonly arises from
the m.3243A$>$G variant in mitochondrial tRNA-Leu
\cite{paper:v6c02-goto-melas-1990}. The clinical presentation is
tissue-specific: the tissues with the highest mitochondrial content
and the highest ATP demand (neurons, cardiomyocytes, skeletal muscle,
kidney) are the first to fail. The same failure mode is induced
acutely by mitochondrial poisons: cyanide blocks Complex IV, rotenone
blocks Complex I, oligomycin blocks ATP synthase. The dose-response
relations for these poisons are textbook examples of the
thermodynamic-limit argument: a cell deprived of oxidative
phosphorylation must rely on anaerobic glycolysis at $2$ ATP per
glucose, and the body's tissues with high ATP demand cannot survive
on the reduced supply.

The genetics of mitochondrial disease introduces the
\engterm{heteroplasmy} concept: each cell contains hundreds to
thousands of mitochondrial DNA copies, and a heterozygous mutation
manifests only when the mutant fraction exceeds a tissue-specific
threshold (commonly $60\,\%$ to $90\,\%$). The threshold is the
fraction at which the remaining wild-type capacity falls below the
tissue's ATP demand; tissues with high ATP demand and low spare
capacity (retinal ganglion cells in LHON, central nervous system in
MELAS) have the lowest thresholds. The engineering reading is that
the body operates with a small reserve margin in its principal
energy supply.

\engterm{The Warburg effect}. Tumour cells, especially under hypoxic
conditions but also under normoxia, sustain a high rate of aerobic
glycolysis with lactate production rather than feeding pyruvate into
the TCA cycle. Otto Warburg's 1956 paper named the effect and
proposed (incorrectly) that mitochondrial dysfunction caused cancer
\cite{paper:v6c02-warburg-1956}. The modern reinterpretation is that
tumour cells reprogramme central metabolism to supply biosynthetic
precursors (NADPH, ribose-5-phosphate, amino acids, fatty acids) at
the cost of ATP-yield efficiency
\cite{paper:v6c02-vander-heiden-2009}. The cell is not failing
oxidative phosphorylation; it is choosing not to use it as the
principal carbon sink. The clinical implication is that
tumour-metabolism-targeting therapies (PI3K inhibitors, glycolytic
inhibitors, IDH-mutant therapeutics) interrupt biosynthesis, not
ATP supply.

\engterm{Insulin resistance and type-2 diabetes}. The body's regulation
of blood glucose is a control system whose actuator is the hormone
insulin, secreted by the pancreatic $\beta$ cells in response to high
glucose. Insulin signals peripheral tissues (muscle, adipose tissue,
liver) to take up glucose and to convert it to glycogen or fat. In
insulin resistance, the peripheral tissues' responsiveness to insulin
is reduced; the $\beta$ cells compensate by secreting more insulin;
chronic over-secretion exhausts the $\beta$ cells; the result is
type-2 diabetes mellitus, characterised by elevated fasting glucose
and progressive complications across multiple organ systems
\cite{paper:v6c02-defronzo-diabetes-2009}. The cellular driver in
many cases is ectopic fat accumulation in muscle and liver, which
interferes with insulin-receptor signalling
\cite{paper:v6c02-shulman-diabetes-2014}. The pathogenesis is, in
engineering terms, a control system whose actuator saturates and
whose set point drifts because the negative-feedback gain has fallen.
The clinical management strategies (diet, exercise, metformin, GLP-1
agonists, insulin therapy) all aim to restore the control-system gain
or to reduce the demand on the actuator. Type-2 diabetes mellitus
affects approximately $500$ million adults worldwide (current as of
2023), making it one of the largest single disease burdens on the
healthcare system.

\engterm{Thermodynamic hype and how to recognise it}. The cellular
metabolic system has been the target of recurring popular claims that
violate the first or second law of thermodynamics: ``perpetual motion
machines'' that pretend to produce more work than they consume,
``zero-point energy'' devices that pretend to extract free energy from
the vacuum, ``free energy'' supplements that pretend to bypass
metabolic limits, ``structured water'' products that pretend to alter
the free energy of intracellular water, ``metabolic typing'' regimens
that pretend to alter substrate fate in defiance of mass balance. The
engineering reading of such claims requires three checks
\cite{text:v6c02-callen-thermodynamics-1985}:

\begin{enumerate}[leftmargin=*]
\item \textbf{First-law check}. Does the proposed device or process
  satisfy energy conservation? An adult human at rest dissipates
  approximately $100\,\text{W}$ of metabolic heat. Any claim of
  reduced food intake without reduced activity, body mass, or body
  temperature must account for the missing energy.
\item \textbf{Second-law check}. Does the proposed device or process
  respect the second law's prohibition on cyclic engines that produce
  net work from a single thermal reservoir? Claims to extract
  ``ambient energy'' from a room-temperature environment without an
  external temperature gradient fail this check.
\item \textbf{Kinetic check}. Does the proposed pathway have a plausible
  enzymatic or catalytic mechanism? A thermodynamically feasible
  reaction is not guaranteed to proceed at any useful rate without
  a catalyst; a claim of room-temperature, uncatalysed CO$_2$ fixation
  to fuels by ``proton resonance'' should be assessed for kinetic
  plausibility.
\end{enumerate}

The three checks together suffice to dispose of most cases. The same
three checks apply to industrial-process claims (the perennial
``free-energy battery'' that absorbs ambient heat to charge itself),
to dietary fads (the ``calorie-free food'' that contains substances
the body cannot metabolise but that, on inspection, are also not food),
and to many proposals in biotechnology and pharmacology. A worked
illustration: a 2010s product line claimed that a proprietary
``activated water'' increased cellular ATP by $30\,\%$ at unchanged
food intake. The first-law check disposes of the claim: the body's
resting heat output ($\sim 100\,\text{W}$) sets the chemical input
rate, and a $30\,\%$ ATP increase at unchanged input requires a
$30\,\%$ reduction in heat output, which would correspond to a body-
temperature drop of several degrees Celsius. No such effect was
measured.

\begin{principle}[Metabolic accounting cannot be cheated]
A cell's free-energy budget is closed: every input has a measurable
source, every output a measurable sink, every flux a thermodynamically
allowed direction. The engineer assessing a metabolic-intervention
claim or a biomanufacturing-yield claim performs the budget by hand
or by FBA, and rejects any claim whose budget does not close. The
budget cannot be cheated because the underlying conservation laws
are exact. Failure modes in this section are failures of the cell's
accounting machinery, not failures of the accounting itself.
\end{principle}

\begin{estimation}
\textbf{Question.} Estimate the daily ATP turnover of an adult human.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
\emph{Method 1: from metabolic rate.} An adult human consumes
approximately $2000\,\text{kcal}/\text{day} = 8.4\,\text{MJ}/\text{day}$
of food energy. Approximately $40\,\%$ is captured as ATP (the body's
overall metabolic efficiency from food to ATP, accounting for $30\,\%$
to $40\,\%$ for substrates entering oxidative phosphorylation and
losses through heat). The captured energy is
$0.4 \times 8.4 = 3.4\,\text{MJ}/\text{day}$. The free energy of ATP
hydrolysis under physiological conditions is approximately
$50\,\text{kJ}/\text{mol}$. The molar ATP turnover is $3.4 \times 10^6 /
5 \times 10^4 = 68\,\text{mol}/\text{day}$, or about $0.78\,\text{mol}/
\text{s}$. The mass turnover (ATP molar mass $507\,\text{g}/\text{mol}$)
is $34\,\text{kg}/\text{day}$.

\emph{Method 2: from cell count and per-cell turnover.} The body
contains approximately $4 \times 10^{13}$ cells
\cite{paper:v6c02-bianconi-cell-count-2013}. The average cell
hydrolyses ATP at approximately $10^7$ molecules per second (rough
average over the body's cell types, weighted toward muscle and
brain). The total ATP hydrolysis rate is $4 \times 10^{20}$ per
second, or $1.4 \times 10^4\,\text{mol}/\text{hour}$, or
$80\,\text{mol}/\text{day}$. The mass turnover is $40\,\text{kg}/\text{day}$.

\emph{Method 3: from oxygen consumption.} An adult human consumes
approximately $250\,\text{mL}/\text{min} = 14\,\text{mol}/\text{day}$
of O$_2$ at rest. Each O$_2$ reduced to water in oxidative
phosphorylation produces approximately $5$ ATP (P/O ratio of $2.5$,
applied to the four electrons per O$_2$ molecule). The ATP
production is $5 \times 14 = 70\,\text{mol}/\text{day}$, or
$36\,\text{kg}/\text{day}$.

\smallskip
\textit{Postmortem.} The three methods give $34$, $40$, and
$36\,\text{kg}/\text{day}$, within $\pm 10\,\%$ of each other. The
textbook value is approximately $50\,\text{kg}/\text{day}$. The
estimate is slightly low; the missing turnover comes from
substrate-level phosphorylation (glycolysis, TCA cycle) that the
oxygen-based estimate does not capture, and from mitochondrial
proton leak (uncoupled respiration that hydrolyses substrates
without producing ATP) that the metabolic-rate estimate does not
subtract.

The remarkable observation is that an organism turns over more than
half its body mass in ATP each day without changing its mass
detectably; the ATP pool is small ($\sim 100\,\text{g}$) and
cycles approximately $500$ times per day.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Three estimates of daily ATP turnover]
The reader produces three independent estimates of an adult human's
daily ATP turnover, by three different methods, and reconciles them
to within a factor of two.

\textbf{Track}. Estimation.
\textbf{Hazard class}: none.

\textbf{Procedure}.

\begin{enumerate}[leftmargin=*]
\item Method A: from a person's daily food energy intake (read the
  Nutrition Facts label of representative foods) and the body's
  metabolic efficiency.
\item Method B: from a per-cell ATP turnover ($10^7$ to $10^9$ molecules
  per second per cell, depending on cell type) and the body's cell
  count.
\item Method C: from a person's oxygen consumption (resting plus
  activity-weighted, with conversion factors from any physiology text)
  and the P/O ratio.
\item Reconcile the three estimates. Identify which assumptions are
  shared across methods (and therefore do not provide independent
  checks) and which are independent.
\end{enumerate}

\textbf{Deliverable}. A five- to eight-page report with three boxed
calculations, a comparison table, and a paragraph on the source of the
largest discrepancy among the methods. Include a sanity-check
calculation of the mass of ATP that would be in the body if the pool
turned over only once per day (and compare to the observed
$\sim 100\,\text{g}$ pool). The reference implementation in
\texttt{code/atp\_turnover\_estimate.py} executes all three methods
and prints a reconciled mean for the reader to compare against.

\textbf{Time}. Three to five hours.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
The standard free energy of ATP hydrolysis is
$\Delta G^{\circ\prime} = -30.5\,\text{kJ}/\text{mol}$. In a cell where
$[\text{ATP}] = 5\,\text{mM}$, $[\text{ADP}] = 0.5\,\text{mM}$,
$[\text{P}_i] = 5\,\text{mM}$, compute the actual free energy of
ATP hydrolysis at $310\,\si{\kelvin}$.
\end{exercise}

\paragraph{Solution sketch.}
Apply $\Delta G = \Delta G^{\circ\prime} + RT \ln Q$ with $Q =
[\text{ADP}][\text{P}_i]/[\text{ATP}] = (5 \times 10^{-4})(5 \times
10^{-3})/(5 \times 10^{-3}) = 5 \times 10^{-4}\,\text{M}$. Then $RT
\ln Q = (8.314)(310)\ln(5 \times 10^{-4}) = (2.58 \times 10^3)(-7.60) =
-19.6\,\text{kJ}/\text{mol}$. Hence $\Delta G = -30.5 - 19.6 \approx
-50\,\text{kJ}/\text{mol}$. The cell drives ATP hydrolysis with
$\sim 50\,\text{kJ}/\text{mol}$ of available work, not the textbook
$30\,\text{kJ}/\text{mol}$.

\begin{exercise}
A cell has $[\text{ATP}] = 5\,\text{mM}$, $[\text{ADP}] = 0.5\,\text{mM}$,
$[\text{AMP}] = 0.05\,\text{mM}$. Compute the energy charge.
\end{exercise}

\paragraph{Solution sketch.}
Plug in: $\text{EC} = ([\text{ATP}] + \tfrac{1}{2}[\text{ADP}])/
([\text{ATP}] + [\text{ADP}] + [\text{AMP}]) = (5 + 0.25)/(5 + 0.5 +
0.05) = 5.25/5.55 = 0.946$. The value sits at the upper end of the
healthy range ($0.85$ to $0.95$); the cell is metabolically rested.
Note that the numerator weights ADP at $\tfrac{1}{2}$ because it
carries half the phosphorylation potential of ATP relative to AMP.

\begin{exercise}
Complete oxidation of palmitate ($\text{C}_{16}\text{H}_{32}\text{O}_2$)
yields $106$ ATP. Compute the ATP yield per gram of palmitate and per
gram of glucose ($32$ ATP). Compare to the energy content per gram
($9.4\,\text{kcal}/\text{g}$ for fat, $4\,\text{kcal}/\text{g}$ for
carbohydrate).
\end{exercise}

\paragraph{Solution sketch.}
Palmitate molar mass $= 256.4\,\text{g}/\text{mol}$, so ATP yield is
$106/256.4 = 0.414\,\text{mol ATP / g}$. Glucose molar mass $=
180.2$, so yield is $32/180.2 = 0.178\,\text{mol ATP / g}$. Ratio
$0.414/0.178 \approx 2.3$; fat yields $\sim 2.3 \times$ more ATP per
gram than carbohydrate. The energy-content ratio in kcal is
$9.4/4 = 2.35$. The match is excellent because both metrics are
proportional to the substrate's reducible content; the small residual
difference comes from the higher carbon-oxidation state of glucose
(some of its energy was already used to oxidise oxygen onto the
sugar). The per-gram comparison is the engineering rationale for
storing chemical energy as fat rather than as glycogen.

\begin{exercise}
Oxidative phosphorylation produces $2.5$ ATP per NADH and $1.5$ ATP
per FADH$_2$. Use the standard reduction potentials of NAD$^+$/NADH
($-0.32\,\text{V}$) and FAD/FADH$_2$ ($-0.22\,\text{V}$ at Complex II
entry) and the O$_2$/H$_2$O couple ($+0.82\,\text{V}$) to compute
the free-energy drop per electron and compare with the per-NADH ATP
yield.
\end{exercise}

\paragraph{Solution sketch.}
Free-energy drop per pair of electrons from NADH to O$_2$ is
$\Delta G = -nF\Delta E^{\circ\prime} = -(2)(96\,485)(0.82 - (-0.32))
= -220\,\text{kJ}/\text{mol NADH}$. At $\Delta G_{\text{ATP}}
\approx -55\,\text{kJ}/\text{mol}$, the maximum thermodynamic ATP
yield per NADH is $220/55 \approx 4$. The cell achieves only $2.5$
ATP per NADH because each transport step dissipates some free energy
as heat and because the proton-pumping stoichiometry is not perfectly
matched. From FADH$_2$ at $E^{\circ\prime} = -0.22$ the drop is
$(2)(96\,485)(0.82 + 0.22) = 200\,\text{kJ}/\text{mol}$, but Complex
II does not pump, so the yield drops to $1.5$ ATP per FADH$_2$. The
$\sim 60\,\%$ efficiency of oxidative phosphorylation is comparable
to a good industrial heat engine.

\begin{exercise}
The Calvin cycle requires three ATP and two NADPH per CO$_2$ fixed.
Compute the photochemical-quantum requirement (photons per CO$_2$)
given that each NADPH requires two photons (one at PSII, one at PSI)
and each ATP requires approximately $1.5$ to $2$ photons through
cyclic and non-cyclic phosphorylation.
\end{exercise}

\paragraph{Solution sketch.}
Two NADPH require $2 \times 2 = 4$ photons. Three ATP at $\sim 1.7$
photons each require $3 \times 1.7 = 5.1$ photons. Total: $9.1$
photons per CO$_2$, conventionally rounded to $10$ in the field. For
$680\,\text{nm}$ red photons at $176\,\text{kJ}/\text{mol}$, the input
free energy is $1760\,\text{kJ}/\text{mol}$ CO$_2$; the stored
free energy (one-sixth of $2870$) is $478\,\text{kJ}/\text{mol}$,
giving a quantum efficiency of $478/1760 = 27\,\%$. Under blue-shifted
illumination ($450\,\text{nm}$, $266\,\text{kJ}/\text{mol}$-photon) the
efficiency drops to $\sim 18\,\%$ because each photon carries more
energy than the photosystem can capture; excess energy is dissipated
as heat at the antenna.

\begin{exercise}
Yeast ferments glucose to ethanol and CO$_2$. Compute the theoretical
maximum mass yield of ethanol per gram of glucose, and verify that
the value $0.511$ matches the stoichiometry $\text{C}_6\text{H}_{12}\text{O}_6
\rightarrow 2\,\text{C}_2\text{H}_5\text{OH} + 2\,\text{CO}_2$.
\end{exercise}

\paragraph{Solution sketch.}
Per mol glucose ($180.2\,\text{g}/\text{mol}$): $2$ mol ethanol
($46.07\,\text{g}/\text{mol}$) and $2$ mol CO$_2$. Mass yield of
ethanol is $(2 \times 46.07)/180.2 = 92.14/180.2 = 0.5113\,\text{g}/
\text{g}$. Mass yield of CO$_2$ is $(2 \times 44.01)/180.2 =
88.02/180.2 = 0.488\,\text{g}/\text{g}$. The sum is $0.5113 + 0.488 =
0.9996 \approx 1$, confirming mass balance to within rounding (the
$0.4\,\%$ residual is the water of hydrolysis released in step 4 of
glycolysis). Industrial yields of $0.45$ to $0.50\,\text{g}/\text{g}$
sit at $88\,\%$ to $98\,\%$ of theoretical; the gap is yeast biomass
and metabolic by-products (glycerol, succinate).

\begin{exercise}
A bioreactor produces $100\,\text{tonnes}$ of ethanol per day from
glucose with $95\,\%$ of theoretical yield. Compute the daily glucose
input in tonnes and the daily CO$_2$ output.
\end{exercise}

\paragraph{Solution sketch.}
At $95\,\%$ of the theoretical $0.5113\,\text{g}$ ethanol per gram
glucose, the actual yield is $0.486\,\text{g}/\text{g}$. Glucose input
is $100/0.486 = 206\,\text{tonnes}/\text{day}$. CO$_2$ output: at the
stoichiometric ratio of $0.488/0.5113 = 0.954$ kg CO$_2$ per kg
ethanol (theoretical) and $95\,\%$ yield giving $0.954 \cdot 0.486 /
0.5113 \approx 0.91$ kg CO$_2$ per kg glucose actually fermented,
hence $206 \cdot 0.91 = 187\,\text{tonnes CO}_2/\text{day}$
(approximately $1.9$ kg CO$_2$ per kg ethanol). A back-of-envelope
cross-check from the simpler ratio $88/46 = 1.91$ kg CO$_2$ per kg
ethanol (pure fermentation stoichiometry) gives $100 \cdot 1.91 =
191\,\text{tonnes CO}_2/\text{day}$, within the rounding bracket.

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the free-energy expression
$\Delta G = \Delta G^{\circ\prime} + RT \ln Q$ for a reaction at
non-standard concentrations from the chemical-potential definition
$\mu_i = \mu_i^\circ + RT \ln a_i$. State the standard-state
conventions used.
\end{exercise}

\paragraph{Solution sketch.}
Write the reaction $\sum_i \nu_i A_i = 0$ with positive $\nu_i$ for
products and negative for reactants. The Gibbs free energy of
reaction at fixed $T, p$ is $\Delta G = \sum_i \nu_i \mu_i$. Substitute
$\mu_i = \mu_i^\circ + RT \ln a_i$: $\Delta G = \sum_i \nu_i \mu_i^\circ
+ RT \sum_i \nu_i \ln a_i = \Delta G^\circ + RT \ln \prod_i a_i^{\nu_i}
= \Delta G^\circ + RT \ln Q$. The $\circ\prime$ convention in
biochemistry is: $T = 298$ or $310\,\text{K}$, $p = 1\,\text{atm}$,
$\text{pH} = 7.0$ (so H$^+$ activity $= 10^{-7}$ is folded into
$\Delta G^{\circ\prime}$), $\text{[H}_2\text{O]}$ activity set to
unity (so water as solvent does not appear in $Q$). For ionic species,
the reference activity is unit molality in an ideal dilute solution.
The derivation is the standard physical-chemistry result
\cite[ch.~3]{text:v6c02-atkins-physical-chem-2018}.

\begin{exercise}
Derive the relationship between the membrane potential $\Delta \Psi$,
the proton-motive force, and the free energy of ATP synthesis at
ATP synthase. Show that the minimum proton-pumping stoichiometry to
synthesise one ATP is set by the ratio of $\Delta G_{\text{ATP}}$ to
the protonmotive free energy $F \Delta \mu_{\text{H}^+}$.
\end{exercise}

\paragraph{Solution sketch.}
The free energy of one mole of protons flowing down the gradient is
$\Delta G_{H^+} = F \Delta \Psi - 2.303 RT \Delta \text{pH}$ (the
sign convention: positive when the cell does work moving protons
\emph{up} the gradient, so the spontaneous flow down has $\Delta G <
0$). At $\Delta \Psi = -150\,\text{mV}$ (matrix-side negative) and
$\Delta \text{pH} = 0.75$, $\Delta G_{H^+} \approx (96\,485)(-0.15)
- (2.303)(8.314)(310)(0.75) = -14.5 - 4.4 = -18.9\,\text{kJ}/\text{mol}$.
The synthase couples $n$ protons in to one ATP out with
$\Delta G_{\text{ATP}} \approx +55\,\text{kJ}/\text{mol}$ (synthesis,
not hydrolysis). Conservation of free energy: $n |\Delta G_{H^+}|
\ge |\Delta G_{\text{ATP}}|$, hence $n_{\min} = 55/18.9 = 2.9$. The
measured value of $\sim 2.7$ protons per ATP
\cite{paper:v6c02-watt-stoichiometry-2010} is consistent with the
thermodynamic minimum (the small shortfall reflects either
mass-action slack or the modern lower estimate of $\Delta G_{\text{ATP}}$).

\begin{exercise}
Derive the steady-state Michaelis-Menten rate law
$v = V_{\max} [S]/(K_M + [S])$ from the rapid-equilibrium assumption
on the enzyme-substrate complex. Discuss when the assumption fails.
\end{exercise}

\paragraph{Solution sketch.}
Mechanism: $E + S \rightleftharpoons ES \rightarrow E + P$ with
forward rate $k_1$, reverse rate $k_{-1}$, and catalytic rate $k_2$.
Assume rapid equilibrium: $[ES] = [E][S]/K_S$ with $K_S = k_{-1}/k_1$.
Conservation: $[E]_0 = [E] + [ES]$. Solve: $[ES] = [E]_0 [S]/(K_S +
[S])$. Rate: $v = k_2 [ES] = k_2 [E]_0 [S]/(K_S + [S]) = V_{\max} [S]/
(K_S + [S])$ with $V_{\max} = k_2 [E]_0$. The Briggs-Haldane
refinement uses the quasi-steady-state assumption $d[ES]/dt = 0$
instead of equilibrium, giving $K_M = (k_{-1} + k_2)/k_1$, which
reduces to $K_S$ when $k_2 \ll k_{-1}$. The rapid-equilibrium
assumption fails when $k_2$ is not small compared to $k_{-1}$, i.e.,
when the catalytic step is not rate-limiting. It also fails in
cellular conditions where $[S]$ is comparable to $[E]_0$ (the total
enzyme is not much less than the substrate), in which case the
free-substrate concentration is not equal to the total substrate
concentration and the substitution $[S] \to [S]_{\text{free}}$ is
required.

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the time required for a typical \emph{E.~coli} cell to
double its mass given a glucose flux of $10^{18}$ glucose molecules
per second, complete oxidation, and a cellular dry mass of
$3 \times 10^{-13}\,\text{g}$.
\end{exercise}

\paragraph{Solution sketch.}
$10^{18}$ glucose/s consumes $10^{18}/6.022 \times 10^{23} = 1.66
\times 10^{-6}\,\text{mol}/\text{s}$. At $180\,\text{g}/\text{mol}$,
$3 \times 10^{-4}\,\text{g}/\text{s}$ of glucose. But this is a flux
for many cells; the question per-cell. Re-read: $10^{18}$ molecules
per second \emph{per cell} would be implausibly large. Assume the
intended scale is the per-cell uptake. \emph{E.~coli} consumes
glucose at $\sim 1\,\text{fmol}/(\text{cell} \cdot \text{s}) =
6 \times 10^{8}$ molecules$/\text{cell}/\text{s}$. Total cellular
glucose-equivalent budget for doubling: $50\,\%$ of dry mass is
carbon, of which $\sim 60\,\%$ comes from glucose, so $\sim 9 \times
10^{-14}\,\text{g}$ glucose per doubling. At $6 \times 10^{8}
\cdot 180/6 \times 10^{23} = 1.8 \times 10^{-13}\,\text{g}/\text{s}$
uptake, doubling time $\approx 9 \times 10^{-14} / 1.8 \times 10^{-13}
\approx 0.5\,\text{s}$. The number is far too short; the
oversimplification is that uptake is not the rate-limiting step.
\emph{E.~coli}'s actual doubling time on glucose minimal medium is
$\sim 60\,\text{minutes}$. The discrepancy ($\sim 7000\times$) is the
combined effect of (a) catabolism diverting most glucose to ATP and
biosynthetic precursors rather than mass accumulation, (b) the
limiting role of ribosome capacity rather than substrate uptake, and
(c) biomass per glucose of $\sim 0.5\,\text{g/g}$ rather than full
mass conversion. The lesson: substrate uptake gives an order-of-
magnitude lower bound on doubling time, which is far from realistic
when biosynthetic machinery is rate-limiting.

\begin{exercise}
Estimate the metabolic cost (in ATP) of synthesising one molecule of
an average protein of mass $40\,\text{kDa}$ ($\sim 360$ amino acids)
from amino-acid precursors, given that each peptide bond costs $4$
ATP equivalents (two for tRNA charging, two for the ribosome cycle).
\end{exercise}

\paragraph{Solution sketch.}
Per peptide bond: $4$ ATP. Per protein: $359 \cdot 4 = 1436\,\text{ATP}$,
rounded to $\sim 1500$. Plus a small correction for proofreading and
elongation factor recycling ($\sim 100$ ATP), total $\sim 1600$.
The energy cost per protein at $50\,\text{kJ}/\text{mol ATP}$ is
$80\,\text{MJ}/\text{mol}$, or $130\,\text{aJ}$ per molecule
(one molecule's free energy of synthesis). At a cellular protein
content of $\sim 2 \times 10^9$ proteins per \emph{E.~coli} and a
doubling time of $\sim 1$ hour, the ATP cost of protein synthesis
alone is $\sim 3 \times 10^{12}\,\text{ATP/cell/hour}$, which
matches the per-cell ATP turnover budget of $\sim 10^7\,\text{ATP/cell/s}$
to within order of magnitude. Protein synthesis is the principal ATP
sink (~25 percent of cellular budget), matching
\cite{paper:v6c02-rolfe-brown-1997}.

\begin{exercise}
Estimate the maximum sustainable mechanical power output of a
$70\,\text{kg}$ athlete given a $\text{VO}_2\text{max}$ of
$70\,\text{mL O}_2/(\text{min}\cdot\text{kg})$, an oxidative-phosphorylation
ATP yield, and a $25\,\%$ mechanical efficiency.
\end{exercise}

\paragraph{Solution sketch.}
Total O$_2$ uptake at max: $70 \cdot 70 = 4900\,\text{mL/min} = 4.9\,
\text{L/min}$. At molar volume $22.4\,\text{L/mol}$: $0.219\,\text{mol
O}_2/\text{min}$, or $0.00365\,\text{mol/s}$. Free-energy yield per
O$_2$ (complete oxidation of glucose released $2870\,\text{kJ}/\text{mol
glucose}$ per $6$ O$_2$): $478\,\text{kJ}/\text{mol O}_2$. Power as
chemical input: $478\,\text{kJ}/\text{mol} \cdot 0.00365\,\text{mol/s}
= 1745\,\text{W}$. Mechanical output at $25\,\%$ efficiency: $1745
\cdot 0.25 \approx 440\,\text{W}$. An elite endurance athlete sustains
$\sim 400\,\text{W}$ for an hour and $\sim 500\,\text{W}$ for a few
minutes; the estimate matches the documented values. The thermal-
load consequence: $1300\,\text{W}$ of waste heat at full output, at
sweat-evaporative cooling of $2400\,\text{kJ/L}$ requires
$\sim 2\,\text{L/h}$ of sweat, which is near the upper sustainable
limit for an acclimatised athlete.

\begin{exercise}
Estimate the photosynthetic biomass productivity in
$\text{g}/(\text{m}^2 \cdot \text{day})$ for an algal
photobioreactor receiving $1\,\text{kWh}/(\text{m}^2 \cdot \text{day})$
of incident light at $5\,\%$ overall efficiency.
\end{exercise}

\paragraph{Solution sketch.}
Captured energy: $1\,\text{kWh/m}^2/\text{day} \cdot 0.05 = 0.05\,
\text{kWh/m}^2/\text{day} = 180\,\text{kJ/m}^2/\text{day}$. Glucose
energy density: $15.6\,\text{kJ/g}$. Biomass yield: $180/15.6 \approx
11.5\,\text{g/m}^2/\text{day}$. Annualised over a $300$-day operational
year: $\sim 35\,\text{t/ha/year}$. This puts the algal reactor between
the C4 crop performance (e.g., maize $\sim 12\,\text{t/ha/year}$) and
the best photovoltaic-then-electrolysis hybrid systems. Comparison to
the values tabulated in \texttt{data/photosynthesis\_efficiency\_by\_crop.csv}:
the $35\,\text{t/ha/year}$ figure matches reported open-pond
\emph{Chlorella} yields and is below closed-photobioreactor performance
under continuous artificial light ($\sim 80\,\text{t/ha/year}$).

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a flux balance analysis of a small toy metabolic network
($10$ reactions, $8$ metabolites) with biomass production as the
objective. Use a linear-programming solver to compute the optimum
flux distribution. Vary the glucose uptake bound and report the
biomass yield curve.
\end{exercise}

\paragraph{Solution sketch.}
The script \texttt{code/metabolic\_flux\_lp.py} solves a
six-reaction, four-metabolite toy version of the network shown in
Figure~\ref{fig:vol06:ch02:mfa-toy}, both analytically and using
\texttt{scipy.optimize.linprog}. Extend the toy: add three more
reactions (e.g., a side branch $C \to E \to \text{biomass}$, a
parallel $A \to B$ path, and an internal cycle); add four more
metabolites. The biomass-yield curve is linear in glucose uptake
until a downstream constraint (e.g., a respiration cap) becomes
binding, at which point the slope drops to zero. The break point
indicates the maximum aerobic capacity. For the \emph{E.~coli}
standard reconstruction the break point is at $\sim 10\,\text{mmol
glucose}/(\text{gDW} \cdot \text{h})$, at which biomass yield is
$\sim 0.5\,\text{gDW/g}$; beyond the break point, additional glucose
is fermented to acetate and the yield drops sharply. Reproduce the
curve with the COBRA toolbox or COBRApy and the iJO1366 model
\cite{paper:v6c02-orth-cobra-2010}.

\begin{exercise}
Simulate a Michaelis-Menten enzyme cascade of three sequential
reactions ($A \rightarrow B \rightarrow C \rightarrow D$) with
specified $K_M$ and $V_{\max}$ for each step. Identify the
rate-limiting step by perturbation analysis and verify by tracking
intermediate concentrations.
\end{exercise}

\paragraph{Solution sketch.}
Integrate the three coupled ODEs $\dot A = -v_1$, $\dot B = v_1 -
v_2$, $\dot C = v_2 - v_3$, $\dot D = v_3$, with $v_i = V_{\max,i}
[\text{substrate}_i]/(K_{M,i} + [\text{substrate}_i])$ and a fixed
initial $A$. At steady state, $v_1 = v_2 = v_3$ (the flux equality is
the definition of steady state in series). The rate-limiting step is
the one with the smallest $V_{\max}$ at the achieved substrate level.
Perturbation: increase each $V_{\max,i}$ by $10\,\%$ in turn; the
step whose perturbation produces the largest steady-state flux
increase is rate-limiting. Verification: intermediates accumulate
\emph{upstream} of the rate-limiting step (the bottleneck). For
example, with $V_{\max} = (10, 1, 10)$ and equal $K_M = 1$, step 2
is rate-limiting, $[B]$ accumulates above $[A]$ and $[C]$, and a
$10\,\%$ increase in $V_{\max,2}$ raises the flux by $\sim 9\,\%$.
The control coefficient $C_i = (\partial v/\partial V_{\max,i})
(V_{\max,i}/v)$ sums to $1$ across the cascade.

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a metabolic-engineering strategy to increase the yield of a
fermentation product. Identify the rate-limiting enzyme of the
biosynthetic pathway, propose an over-expression strategy, and
estimate the maximum theoretical yield from the stoichiometry. State
the secondary constraints that may bottleneck the actual yield
(cofactor balance, redox state, by-product inhibition).
\end{exercise}

\paragraph{Discussion.}
A rubric for the strategy: (1) draw the full pathway from substrate
to product, including all branch points and cofactor exchanges; (2)
identify the rate-limiting step by FBA flux variability analysis or
$^{13}$C-MFA on the wild-type strain; (3) propose an over-expression
of the rate-limiting enzyme using a strong constitutive promoter or
a copy-number-amplified plasmid; (4) compute the maximum theoretical
yield from substrate-product stoichiometry ignoring biomass costs;
(5) list the cofactor balance constraint (does the pathway consume
or produce excess NADPH? does it require ATP? does it generate
acetyl-CoA out of stoichiometric ratio with the product?); (6)
identify by-product inhibition risks (acetate at high glucose flux
in \emph{E.~coli}, ethanol in yeast). A concrete example: succinate
production from glucose has a theoretical maximum of $1.71\,\text{mol
succ/mol glc}$ via the reductive TCA branch; actual industrial yields
are $\sim 1.1\,\text{mol/mol}$, the gap being acetate by-production
and biomass diversion. Good answers explicitly quantify the gap.

\begin{exercise}
Design a respiratory inhibitor that targets a specific complex of the
mitochondrial electron transport chain for use as a research tool.
Identify the molecular feature of the chosen complex that distinguishes
it from the others and the safety considerations for the design.
\end{exercise}

\paragraph{Discussion.}
A working rubric: (1) name the target complex and the unique
structural feature (Complex I's quinone-binding site; Complex II's
flavin domain; Complex III's $Q_o$ pocket; Complex IV's binuclear
copper-haem centre; ATP synthase's $F_0$ rotor); (2) propose a small-
molecule scaffold known to bind the feature (rotenone for Complex I;
malonate for Complex II; antimycin or stigmatellin for Complex III;
cyanide or carbon monoxide for Complex IV; oligomycin for ATP
synthase $F_0$); (3) discuss isoform selectivity (is the inhibitor
also active against bacterial homologues? plant homologues? if used
in animal studies, what is the cardiac vs neuronal preference?);
(4) discuss the safety considerations (acute toxicity of Complex IV
inhibitors at minute doses, the LD$_{50}$ of $\sim 1.5\,\text{mg/kg}$
oral for cyanide makes it an unsuitable research reagent except in
isolated-mitochondria preparations; rotenone and oligomycin are
preferred for whole-cell work). Strong answers cite specific binding
constants (e.g., rotenone's $K_i \approx 4\,\text{nM}$ on bovine
Complex I) and propose handling protocols matched to the molecule's
toxicity.

\begin{exercise}
Design a feedback-control strategy for a bioreactor that maintains
the substrate concentration at a setpoint by adjusting the substrate
feed rate based on the dissolved oxygen reading. Specify the
controller type and the expected response time.
\end{exercise}

\paragraph{Discussion.}
A rubric: (1) draw the control loop: dissolved-O$_2$ sensor $\to$
controller $\to$ substrate feed pump; (2) note that the indirect
inference relies on the Monod relation between substrate excess and
oxygen demand, which means the loop is implicit and has a phase lag
of seconds (the cell's response to substrate change) plus seconds
(the sensor's response time); (3) propose a PI controller with
integral time $\sim 30\,\text{s}$ and proportional gain tuned for a
critically damped response; (4) note the failure modes: at very low
oxygen (the loop saturates because cells are oxygen-limited rather
than substrate-limited) and at very high oxygen (the loop is too
insensitive). Strong answers add a feedforward term from the known
glucose-uptake-vs-OD relation and a hard upper bound on feed rate to
prevent overfeeding. Industrial bioreactors typically run a
dissolved-O$_2$-based feedback loop with a stirrer-speed cascade.

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A patient with chronic muscle weakness has elevated serum lactate at
rest, normal at exertion. Identify which metabolic compartment is
likely to be impaired and propose one diagnostic measurement that
would distinguish a glycolytic enzyme deficiency from a mitochondrial
respiratory-chain deficiency.
\end{exercise}

\paragraph{Solution sketch.}
Elevated lactate at rest with normal-during-exertion is the inverse
of the classical pattern (rest-normal, exertion-high), and points to
a chronically suppressed mitochondrial respiratory chain that the
cell compensates for with anaerobic glycolysis at rest; during
exertion the patient cannot increase lactate further because
glycolytic capacity is already saturated. The likely compartment is
mitochondrial (a respiratory-chain or coenzyme-Q defect), not
glycolytic. Diagnostic to distinguish: measure the lactate/pyruvate
ratio. A glycolytic enzyme deficiency lowers both lactate and
pyruvate (the substrate of LDH is depleted), so the ratio remains
near $10:1$. A mitochondrial defect raises NADH/NAD$^+$ in the
cytosol, which drives LDH toward lactate, so the ratio rises to
$\sim 30:1$ or higher. A muscle biopsy with respirometry on
permeabilised fibres is the gold-standard follow-up.

\begin{exercise}
A bacterial culture growing on glucose accumulates acetate in the
medium even when oxygen is plentiful. Identify the metabolic state
(overflow metabolism) and the likely rate-limiting step in the
respiratory chain. Propose a metabolic-engineering intervention that
would reduce acetate accumulation.
\end{exercise}

\paragraph{Solution sketch.}
The phenomenon is the bacterial Crabtree effect, the prokaryotic
analogue of the Warburg effect in tumours. Acetate accumulates
because glycolytic flux at high glucose uptake exceeds the TCA-cycle
and respiratory-chain capacity; the excess acetyl-CoA is diverted to
acetate via phosphotransacetylase and acetate kinase. The rate-
limiting step in the respiratory chain is typically the membrane
density of cytochromes (the cell cannot fit more respiratory
machinery in the inner membrane). Interventions: (1) attenuate
glucose uptake via promoter engineering on \texttt{ptsG} (the
PTS-system transporter); (2) overexpress NADH dehydrogenase or
fumarase to increase TCA capacity; (3) delete the \texttt{pta} or
\texttt{ackA} genes (the acetate-excretion path), forcing carbon
back through the TCA cycle. Approach (1) gives the largest practical
gain in \emph{E.~coli} fed-batch fermentations.

\begin{exercise}
A microalgal culture's photosynthetic yield drops sharply when light
intensity exceeds a threshold. Identify the engineering mechanism
(photoinhibition: damage to the PSII reaction centre at high photon
flux) and propose two reactor-design strategies (light-path geometry,
shading) that would prevent the saturation.
\end{exercise}

\paragraph{Solution sketch.}
Photoinhibition arises when the rate of PSII reaction-centre damage
exceeds the rate of D1-protein repair; the steady-state PSII
abundance falls, and the photosynthetic rate drops. The threshold
photon flux at which photoinhibition becomes net negative is
$\sim 500\,\mu\text{mol photons}/(\text{m}^2 \cdot \text{s})$ for most
mesophilic microalgae, well below midday sunlight ($\sim 2000\,\mu
\text{mol}/(\text{m}^2 \cdot \text{s})$). Two reactor-design responses:
(1) thin-film flat panels with optical-path lengths of $1$ to
$2\,\text{cm}$, which distribute the photon flux across more cells
and reduce the per-cell instantaneous exposure (the same total
photon budget is spread thinner); (2) light-diluting shading
materials, e.g., wavelength-selective films that filter UV and
near-IR while transmitting PAR, or vertically-oriented bag
photobioreactors that catch oblique morning and evening light at
higher angle. Industrial outdoor algal raceways typically operate at
$\sim 30\,\text{cm}$ depth and rely on self-shading by the algal
biomass itself; closed photobioreactors use light-path tuning more
aggressively.

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A worker exposed to cyanide gas loses consciousness within minutes.
Identify the molecular target (cytochrome $c$ oxidase, Complex IV),
trace the consequence to ATP synthesis, and identify the tissue
(brain or heart) that fails first.
\end{exercise}

\paragraph{Solution sketch.}
Cyanide binds the haem iron in the binuclear centre of cytochrome
$c$ oxidase (Complex IV) with high affinity ($K_d \sim 10^{-4}\,
\text{M}$ but kinetically rapid), blocking the four-electron
reduction of O$_2$ to water. The electron transport chain backs up
(all upstream carriers become reduced); proton pumping stops;
$\Delta \mu_{H^+}$ collapses; ATP synthase reverses (hydrolyses ATP
to maintain the gradient, depleting the cellular ATP pool). The
brain fails first because it has the highest ATP demand per unit
mass and the smallest energy reserve (no significant glycogen, no
fat); within $\sim 10$ seconds, cortical neurons lose function and
consciousness is lost. The heart follows within minutes if not
rescued. The clinical antidote (hydroxocobalamin or
sodium thiosulfate plus sodium nitrite) sequesters cyanide as
cyanocobalamin or thiocyanate; restoration of Complex IV activity
permits recovery within minutes if administered before irreversible
neuronal damage.

\begin{exercise}
A patient with severe sepsis has elevated lactate, low blood pressure,
and tissue hypoxia despite normal arterial oxygen saturation. Identify
the cellular metabolic state (uncoupled mitochondrial respiration or
impaired oxygen extraction) and the engineering mechanism by which
the elevated lactate develops.
\end{exercise}

\paragraph{Solution sketch.}
Septic shock produces a paradoxical state in which arterial blood is
fully oxygenated yet tissues are hypoxic; the failure is in tissue
oxygen extraction, not delivery. Two mechanisms contribute: (1)
microvascular shunting, in which capillary perfusion is heterogeneous
and some capillary beds are bypassed entirely; (2) mitochondrial
dysfunction, in which inducible nitric oxide synthase (iNOS) generates
NO that inhibits Complex IV and uncouples respiration. The cellular
consequence is the same as cyanide intoxication, but slower: the
NADH/NAD$^+$ ratio rises, LDH drives pyruvate to lactate, and lactate
spills into the blood. The metabolic signature (high lactate, normal
to high arterial $\text{SaO}_2$, low arteriovenous oxygen difference)
distinguishes septic shock from haemorrhagic shock (high lactate but
low $\text{SaO}_2$). Treatment targets perfusion (fluids,
vasopressors) and reverses the mitochondrial dysfunction over hours
to days; mortality remains $\sim 30\,\%$ once shock develops.

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A health supplement advertises an ``ATP-boosting'' formula that
allegedly increases cellular ATP levels by $30\,\%$. Identify the
engineering checks (first-law, second-law, kinetic) that the claim
must satisfy and write a one-paragraph reply identifying the likely
mode of failure of the claim.
\end{exercise}

\paragraph{Discussion.}
A strong reply notes: cellular ATP concentration is held at $\sim
5\,\text{mM}$ by the homeostatic balance between synthesis (oxidative
phosphorylation + glycolysis) and demand. The pool size is small
relative to turnover ($\sim 1$ to $2\,\text{s}$ half-life), so a
sustained $30\,\%$ increase would require a $30\,\%$ increase in
either synthesis or a $30\,\%$ decrease in demand, neither of which
is plausible without major intervention. First law: if synthesis rose
by $30\,\%$, food intake must rise commensurately ($\sim 600\,\text{kcal/
day}$) or substrate must come from body stores (visible weight
loss). Second law: no claim that the supplement reduces metabolic
heat output is offered, and would be detectable by infrared
thermography. Kinetic: no plausible enzyme or transporter is named
that the supplement modulates; ``ATP-boosting'' is not a defined
biochemical action. The likely mode of failure is that the
supplement does nothing measurable; the placebo effect on subjective
energy may produce positive customer testimonials.

\begin{exercise}
A start-up claims to produce hydrogen fuel directly from atmospheric
CO$_2$ using ``optimised photosynthetic bacteria'' at $20\,\%$
overall solar-to-hydrogen efficiency. Compare the claim to the
thermodynamic maximum efficiency of photosynthesis and to the
best-measured laboratory yields of bacterial photohydrogen
production. Write a one-paragraph reply identifying the assumption
that the claim is most likely violating.
\end{exercise}

\paragraph{Discussion.}
Photosynthesis caps at $\sim 27\,\%$ quantum efficiency at $680\,\text{nm}$
(photons-to-stored-glucose, accounting for the $10$-photon-per-CO$_2$
requirement) and at $\sim 12\,\%$ whole-spectrum-to-stored-glucose
(PAR is only $\sim 45\,\%$ of incident solar). Best-documented
laboratory bacterial photohydrogen yields are $\sim 1\,\%$ to $3\,\%$
solar-to-H$_2$ \cite{paper:v6c02-zhu-photosynthesis-2010}. A $20\,\%$
claim exceeds the thermodynamic photosynthesis maximum even before
accounting for the inefficiency of routing carbon to H$_2$ rather
than to biomass. The most likely violated assumption is that the
quoted efficiency is per-spectral-window (e.g., only the PSII-PSI
absorption band of $400$ to $700\,\text{nm}$ is counted as input
solar) rather than per-incident-spectrum (which is the engineering-
relevant metric for outdoor deployment). Strong answers also note
that the claim implicitly assumes complete cellular carbon channelling
to H$_2$, which leaves no carbon for biomass and is incompatible with
sustained cell viability.

\begin{exercise}
A textbook describes mitochondria as ``the powerhouses of the cell.''
Identify two ways in which the powerhouse metaphor is accurate
(centralised energy conversion, distributed delivery) and two ways in
which it is misleading (the cell does not transmit ATP as electrical
power, and mitochondria perform many non-energetic roles).
\end{exercise}

\paragraph{Discussion.}
Accurate: (1) mitochondria account for $\sim 90\,\%$ of cellular ATP
synthesis under aerobic conditions, so they are the principal energy
converter, analogous to a centralised power station; (2) ATP is
delivered throughout the cell by diffusion and by the
adenylate-translocator family, fulfilling the ``distributed delivery''
role analogous to an electrical grid. Misleading: (1) the cell does
not transmit power as voltage and current; ATP is a chemical
intermediate that must be hydrolysed at the point of use, so the
analogy fails on the medium of transport; (2) mitochondria are the
site of multiple non-energetic processes (apoptosis initiation,
calcium buffering, iron-sulfur cluster biosynthesis, heme synthesis,
steroidogenesis in adrenocortical cells), so describing them solely
as power-conversion units understates their function; (3) the
mitochondrial network is highly dynamic (fission and fusion, with
turnover by mitophagy on the order of weeks), unlike a static power
plant; (4) different tissues have radically different mitochondrial
densities (cardiomyocytes $\sim 35\,\%$ cell volume vs.\ adipocytes
$\sim 5\,\%$), reflecting tissue-specific energy demand rather than a
single centralised model.

\begin{exercise}
A diet popular at the time of writing (current as of 2025) claims
that ``calories don't count'' for certain food types. Identify the
first-law claim being made and propose a controlled study design that
would test it.
\end{exercise}

\paragraph{Discussion.}
The first-law claim is that the food contains caloric energy that
the body cannot metabolise, so the body's energy uptake is less
than the food's bomb-calorimeter energy. This is true for some food
constituents (dietary fibre, resistant starches, sugar alcohols),
which the body partially excretes or routes to colonic bacteria;
the body's metabolisable energy of fibre is $\sim 2\,\text{kcal/g}$,
not the $4\,\text{kcal/g}$ that a bomb calorimeter would report. The
plausible scientific reading is therefore that the diet exploits a
known calibration gap, not that calories do not count. A controlled
study design: a randomised crossover trial in which participants
consume isocaloric (by bomb-calorimeter measure) diets composed of
the claimed-special food vs.\ a standard control, with weight, body
composition (DEXA), and resting metabolic rate (indirect calorimetry)
measured weekly. Stool calorimetry quantifies the unabsorbed fraction
directly. The metabolisable-energy gap, if it exists, is bounded by
the stool-calorimetry difference. The trial requires $\sim 20$
participants per arm and a $4$-week duration to detect a $\sim 5\,\%$
metabolisable-energy difference. The claim, if true at all, is a
quantitative correction, not a violation of the first law.

\begin{exercise}
A start-up reports a microbial process that converts methane to
liquid hydrocarbons at $90\,\%$ carbon yield without external energy
input. Identify the thermodynamic check that disqualifies the claim
and the kinetic check that would be needed even if the thermodynamics
allowed it.
\end{exercise}

\paragraph{Discussion.}
The thermodynamic check: complete combustion of methane releases
$\sim 890\,\text{kJ/mol CH}_4$; complete combustion of liquid
hydrocarbons (e.g., n-octane) releases $\sim 5470\,\text{kJ/mol C}_8\text{H}_{18}$,
or $\sim 684\,\text{kJ/mol carbon}$. Per carbon, n-octane releases
slightly less energy than methane (because methane is more reduced
per carbon), so converting methane to a liquid alkane is roughly
isothermal at the carbon level. The thermodynamic check therefore
\emph{does not} disqualify a $90\,\%$ carbon yield outright; the
process is approximately energy-neutral if the products are kept in
the reduced state. The check that disqualifies the claim depends on
the specific product: if the product is a partial-oxidation species
(methanol, formic acid), the process is exergonic and feasible; if
the product is fully reduced (alkane chains), the process requires
a net reductant (H$_2$ or e$^-$), and ``without external energy
input'' fails. The kinetic check: methane is among the most
chemically inert of organic molecules ($\text{C-H}$ bond energy
$\sim 440\,\text{kJ/mol}$). Any enzymatic or catalytic process that
activates methane at room temperature must be specified by name and
its turnover number measured. The only known biological methane
activator is the methane monooxygenase enzyme of methanotrophic
bacteria, whose $k_{\text{cat}} \sim 1\,\text{s}^{-1}$ limits the
volumetric productivity to $\sim 1\,\text{g/L/day}$ at maximum cell
density, which is far below commodity-fuel economic thresholds.

\begin{exercise}
A colleague proposes that ``brown fat'' tissue, which dissipates
metabolic energy as heat without producing ATP, violates the
free-energy efficiency of oxidative phosphorylation. Identify the
engineering mechanism (proton leak via uncoupling protein UCP1) and
explain why the dissipation is regulated, not anomalous. Write a
one-paragraph reply situating brown fat in the body's thermal-control
system.
\end{exercise}

\paragraph{Discussion.}
Brown adipose tissue expresses uncoupling protein 1 (UCP1), a proton
channel in the mitochondrial inner membrane that short-circuits the
proton gradient, dissipating its free energy directly as heat rather
than routing it through ATP synthase
\cite{paper:v6c02-cannon-brown-fat-2004}. The dissipation does not
violate any thermodynamic principle: it converts the chemical free
energy of substrate oxidation directly into heat, which is the
maximally degraded form of energy but is also the most thermodynamically
allowed. The proton leak is regulated: UCP1 is activated by free
fatty acids and inhibited by purine nucleotides (ATP, ADP, GTP, GDP),
and the body adjusts UCP1 expression and activity in response to cold
exposure. Brown fat is therefore a regulated heat generator, part of
the body's thermal-control system, with $\sim 50\,\text{g}$ of active
brown fat in an adult capable of generating $\sim 100\,\text{W}$ of
heat under maximal stimulation, comparable to a small space heater.
The body trades $40\,\%$-efficient ATP synthesis for $100\,\%$-efficient
heat production during cold exposure. The trade is energetically
expensive but evolutionarily favoured in small mammals and in human
infants. See also \cite{paper:v6c02-jastroch-uncoupling-2010} on the
broader role of basal proton leak in setting standard metabolic rate.
