\chapter{The capstone problem}
\label{vol11:ch13}

\epigraph{Volumes I through XI taught the reader to think and to design like an engineer; the capstone asks the reader to act like one.}{}

\chapmeta{Half-life: medium-life. Archetypes: ethics. Exercise target: 0. Project track: hybrid.}

\noindent
The capstone is the work the reader does to integrate the full discipline of this book into a single engineering project of meaningful scale. The chapters of Volume XI taught the discipline element by element; the studios of Chapter 12 integrated the elements at depth on bounded problems; the capstone is the reader's choice of a problem at the scale of professional engineering and the application of the integrated discipline to it. The capstone is the engineering education's culmination; the chapter's purpose is to frame what the capstone demands and what its successful completion represents.

\section{The capstone framing}\pathtag{core}

The capstone is the moment the book stops asking the reader to learn and starts asking the reader to act. Volumes I through X taught the reader to think like an engineer: the mathematics, the materials, the failures, the systems, the disciplines. Volume XI taught the reader to design like an engineer: the requirements, the concepts, the details, the analyses, the prototypes, the tests, the reviews, the ethics. The capstone asks the reader to do what an engineer does: take a real problem at meaningful scale, apply the full discipline, produce a deliverable that a competent professional would recognise as engineering work.

The capstone differs from the chapter projects and from the studios in scale and substance:

\textbf{Scale.} The chapter projects are scoped to days or weeks; the studios are scoped to months. The capstone is scoped to the time required to do substantive engineering work: typically six months to a year of part-time effort, or several months of concentrated effort.

\textbf{Substance.} The chapter projects develop specific skills; the studios integrate skills on bounded problems; the capstone is a problem the reader chose because it is genuinely worth solving. The reader's commitment to the problem is the capstone's substantive driver.

\textbf{Stakes.} The chapter projects and the studios are exercises; the capstone is intended to produce a deliverable the reader could defend to a professional design panel. The deliverable should be substantial enough that the engineering judgment it required is real engineering judgment.

\textbf{Accountability.} The chapter projects and studios are between the reader and the book; the capstone is between the reader and the broader engineering community. The reader's choice of problem, the choices made in solving it, and the resulting design pack are professional commitments.

The capstone is not a graduation exercise in the academic sense; this book does not confer credentials. The capstone is the engineering work the reader does to confirm to themselves and to those they show the work to that the reader has integrated the discipline this book teaches. The confirmation is the substantive value of the capstone, not any external certification.

\section{The capstone brief: a real engineering problem at meaningful scale}\pathtag{core}

The capstone problem must be real, meaningful, and at scale. The criteria:

\textbf{Real.} The problem exists in the world; people are affected by it; the design's success or failure has consequences beyond the reader's exercise. Real problems differ from textbook problems in that their requirements are not pre-specified, their constraints are not pre-listed, their evaluation criteria are not pre-defined; the engineer must engage with all of these.

\textbf{Meaningful.} The problem is worth solving. Meaningful in the engineering sense: the design's deployment would address a need that the current situation does not adequately address. The reader chooses the problem partly on the basis of personal commitment; the reader will spend months on it; the meaning sustains the commitment.

\textbf{At scale.} The problem is large enough that the engineering discipline is exercised at substantive depth. A problem that can be solved in days is too small; a problem that cannot be solved by the reader alone is too large. The scale is calibrated to the reader's situation: part-time work over months, with realistic resources and access.

The categories of capstone problems that this book has prepared the reader to take on:

\textbf{A small bridge for a community.} A pedestrian bridge across a small stream serving a community whose current crossing is inadequate. The problem engages: structural design (Volume III); materials selection (Volume II); environmental considerations (Volumes V and X); accessibility (Volume XI Chapter 10); community engagement (Volume XI Chapter 3); regulatory framework (local building codes); ethical considerations (the community's relationship with the design and with the design's outcome).

\textbf{A community-scale water-purification system.} A water-treatment system serving 100-500 people in a community without reliable potable-water access. The problem engages: chemistry (Volume V); fluid mechanics (Volume III); biological considerations (Volume VI); appropriate-technology design (Volume XI Chapter 12 Studio 1); maintenance and lifecycle (Volume XI Chapter 10); ethics of intervention in a community.

\textbf{A small-batch manufacturing line.} A production line for an artefact the reader has prototyped (perhaps from Chapter 7's prototype project). The problem engages: manufacturing-process design (Volume VIII); quality systems (Volume X); workforce considerations (Volume XI Chapter 11); economic feasibility; supply-chain design; environmental footprint.

\textbf{An open-source medical device.} A medical device intended for use in settings where commercial devices are unavailable or unaffordable. The problem engages: safety-critical software (Volume XI Chapter 12 Studio 4); medical-device regulatory framework (Volume X Chapter 11); user research with clinical populations; ethics under risk; the open-source distribution model.

\textbf{A robotic aid for a disability.} An assistive device for users with specific disabilities the reader can engage with. The problem engages: human-centred design (Volume XI Chapter 3); accessibility (Volume XI Chapter 10); mechanical design (Volume III, Volume VIII); software (Volume VII); ethics of designing for users different from the designer.

\textbf{A renewable installation.} A specific renewable-energy installation for a specific site (the reader's home; a community building; a small business). The problem engages: energy engineering (Volume IV); electrical engineering (Volume VIII); structural design for mounting (Volume III); the studio approach of Chapter 12 Studio 3; economic and environmental analysis.

\textbf{A public-data platform with safety implications.} A software platform that collects and analyses data with implications for individuals or for public policy. The problem engages: software architecture (Volume VII); data engineering; safety considerations (Volume X Chapter 5); privacy and ethics (Volume XI Chapter 11); regulatory framework for data protection.

The categories are illustrative, not exhaustive. The reader chooses a problem that fits the reader's circumstances, interests, and access. The choice is itself a substantive design decision; it shapes everything that follows.

\section{What the reader brings: every previous volume}\pathtag{standard}

The capstone is the integration of the entire book. The reader brings:

\textbf{From Volume I (Quantity).} The discipline of measurement: how to characterise the quantities the design depends on; how to bound the uncertainties; how to make estimation a working tool rather than guesswork.

\textbf{From Volume II (Form).} The understanding of geometry and structure: how shapes interact with forces; how form choices constrain the design space; how tolerances accumulate.

\textbf{From Volume III (Force).} The mechanics of solids and structures: how loads produce stresses; how structures resist failure; how the analytical tools (closed-form; FEA) support detailed design.

\textbf{From Volume IV (Energy).} The thermodynamics, fluid mechanics, and energy systems: how energy flows constrain the design; how efficiency considerations shape choices; how the energy budget is the dominant lifecycle consideration for many designs.

\textbf{From Volume V (Matter).} The materials engineering: how material properties shape the design space; how material selection trades off across multiple criteria; how processing affects properties.

\textbf{From Volume VI (Life).} The biological-engineering disciplines: how biology can be engineering subject; how living systems interface with engineered ones; how the biological context constrains certain designs.

\textbf{From Volume VII (Information).} The computational and informational engineering: how information flows shape modern designs; how software discipline applies to engineering software; how the data systems the design depends on are themselves designed artefacts.

\textbf{From Volume VIII (Machines).} The mechanical and electromechanical machines: how the design's moving parts work; how the mechanisms, the electronics, the actuators, the sensors integrate.

\textbf{From Volume IX (Systems).} The systems-engineering discipline: how subsystems compose into systems; how control loops manage system behaviour; how the integration is the substantive engineering.

\textbf{From Volume X (Failure).} The failure literacy: how systems fail; what the failure modes are across the disciplines; how the discipline of designing-against-failure shapes the substantive engineering work.

\textbf{From Volume XI (Design), Chapters 1-12.} The integrated design discipline: how the requirements become specifications, the concepts become detailed designs, the analyses confirm or refute the design's adequacy, the prototypes surface what the analyses miss, the tests produce the evidence, the reviews close the loop, the lifecycle considerations shape the design from the start, and the ethics frames every decision.

The capstone is where these come together. No single chapter and no single volume produces the capstone's substance; the integration is what makes the capstone the engineering work the volumes prepared for.

\section{The capstone deliverable: a design pack reviewable by a panel}\pathtag{core}

The capstone's deliverable is a complete design pack. The pack should be substantial enough that a competent professional design panel could review it and arrive at a substantive judgment of the design's adequacy. The pack's contents:

\textbf{Problem statement.} What problem the design addresses; who is affected; what the existing situation is; what need the design serves.

\textbf{Stakeholders and user research.} Who the design's stakeholders are; what user research was conducted; what findings emerged; how the findings informed the design.

\textbf{Requirements specification.} The functional and non-functional requirements; the priorities; the acceptance criteria; the verification approach.

\textbf{Concept generation and selection.} The concepts that were generated; the evaluation against requirements; the selected concept; the rationale.

\textbf{Detailed design.} The drawings, models, or specifications that constitute the design; the materials and components; the manufacturing or construction approach; the tolerances and specifications.

\textbf{Analyses.} The structural, thermal, electrical, control, lifecycle, or other analyses that support the design's adequacy claims; the assumptions; the validation against benchmarks where possible; the uncertainty characterisation.

\textbf{Simulation evidence.} Where simulation supports the design, the verification and validation of the simulations; the results; the design implications.

\textbf{Prototype evidence.} The prototypes built; what they tested; what they showed; how they informed the design.

\textbf{Test plan and results.} The verification plan; the tests conducted; the results; the conclusions against acceptance criteria.

\textbf{Review records.} The reviews conducted; the comments raised; the resolutions; the institutional commitment the reviews represent.

\textbf{Lifecycle considerations.} Manufacturability; maintainability; sustainability; accessibility. The trade-offs made and the rationales.

\textbf{Failure-mode analysis.} The failure modes identified; the mitigations; the residual risks; the operational considerations.

\textbf{Ethical considerations.} The risk acceptance; the populations served and excluded; the disclosure obligations; the dual-use considerations; the reader's professional commitment to the design.

\textbf{Documentation for implementation.} The drawings, work instructions, parts lists, service manuals, training materials that would support actual deployment.

\textbf{Reflection.} What the reader learned from the capstone; what would be done differently next time; what the design's relationship is to the engineering profession's broader discipline.

The pack is substantial: tens of pages for simpler projects; hundreds for complex ones. The substance is in the engineering work the pack documents, not in the pack's size. A short pack that documents substantive engineering is better than a long pack that elaborates without depth.

The pack should be reviewable: a competent professional should be able to read it and form a judgment about whether the design is adequate. The reader's discipline in producing the pack is the discipline of producing engineering documentation; the documentation is itself part of the engineering work.

\section{What graduation looks like}\pathtag{standard}

The reader who completes the capstone has achieved the integration this book is designed to support. The graduation is not a credential but a working state: the reader can do the work of an engineer at meaningful scale, with the discipline this book teaches, accountable to the engineering profession's standards.

What this looks like in practice:

\textbf{The reader can take on a substantive engineering problem.} Not by reciting techniques but by exercising the integrated discipline: bounding the problem; engaging stakeholders; generating and evaluating concepts; producing detailed designs; analysing and testing; reviewing and signing; considering the lifecycle; reflecting ethically.

\textbf{The reader has working judgment.} The reader has internalised when to apply which techniques, how deep to go in each, when to commit and when to iterate. The judgment is what the chapters could not teach directly; the studios and the capstone develop it through practice.

\textbf{The reader has a professional identity.} The reader thinks of themselves as an engineer in the substantive sense developed by this book: someone whose work is the discipline of producing reliable intervention under constraint, failure, scale, and responsibility. The identity shapes the reader's subsequent engagement with engineering work, regardless of formal credentials.

\textbf{The reader has a defensible body of work.} The capstone is the central piece; the chapter projects and the studio (if completed) are supporting evidence; the reader can show the work to colleagues, prospective employers, professional bodies, and the engineering community. The body of work is the reader's professional portfolio.

\textbf{The reader has a calibrated sense of their limits.} The discipline this book teaches includes the discipline of knowing what the reader does not know. The reader can identify the problems within their competence and the problems that require additional expertise; the reader knows when to consult, when to defer, when to escalate.

\textbf{The reader has a continuing commitment.} The capstone is not the end of the reader's engineering development. The reader has the discipline that supports continued development: through professional engagement; through continued reading; through future engineering work; through teaching and mentoring others.

The reader who completes the capstone has, by the book's measure, become an engineer in the substantive sense. The formal recognition of the achievement is separate: licensure where applicable; professional membership; academic credentials; employment in engineering positions. The substantive engineering capability is what the book prepares; the formal recognitions are sought through other channels.

\section{The reader's place after the book}\pathtag{standard}

The book is the formation; the reader's career is the application. The book's discipline carries forward into the reader's subsequent engineering work in several directions:

\textbf{Continued practice.} The reader's engineering work is the application of the discipline at scale. The reader takes on engineering problems; produces designs; engages with the engineering community; contributes to the profession's collective work. The substantive engineering practice is the continuing expression of the book's discipline.

\textbf{Continued learning.} The engineering disciplines evolve; the reader's competence is maintained through continued learning. New techniques emerge; new failure modes are documented; new ethical questions arise; new technologies enable new approaches. The reader's engagement with the engineering literature, with the professional bodies, with peer engagers is the continuing learning.

\textbf{Teaching and mentoring.} The reader contributes to other engineers' formation through teaching (formally or informally), through mentoring junior colleagues, through writing, through participating in professional bodies. The reader's contribution to the profession's continuation is partly through the next generation's development.

\textbf{Institutional contribution.} The reader contributes to the institutions that the engineering profession depends on: the standards bodies, the regulatory frameworks, the professional bodies, the educational institutions. The institutional work is the engineering profession's collective self-construction; the reader's participation is part of the profession's continued existence.

\textbf{Public engagement.} The reader contributes to the public's understanding of engineering: through clear communication about engineering decisions affecting the public; through participation in public discussions of technology, infrastructure, and environment; through the substantive engagement that distinguishes engineering's public voice from marketing or advocacy.

\textbf{Volume XII engagement.} The book's twelfth volume addresses engineering at civilization scale: the planetary, infrastructural, intergenerational, and political dimensions of engineering practice. The capstone's discipline is the precondition for engaging with Volume XII's domain; the reader who has integrated the discipline through the capstone is prepared to consider engineering's responsibilities at scales beyond individual projects.

\section{Closing: handing the reader to Volume XII}\pathtag{core}

This volume has been the volume of design: the discipline by which the engineer takes a need and produces an artefact. The volume's eleven previous chapters developed the discipline element by element; the studios of Chapter 12 integrated the elements; this chapter has set up the capstone as the reader's integrated practice.

The book's arc is now near its end. Volumes I through X built the foundations; Volume XI built the integrated design discipline; the capstone is the reader's working application. Volume XII addresses what is left: engineering at civilization scale. The engineering profession is responsible not only for individual projects (this volume's scope) but for the cumulative effect of engineering work on civilisation: the planet's climate; the infrastructure's continuation; the technological framework within which subsequent generations will operate; the engineering knowledge that is built up and that is lost. Volume XII is the volume's address of these.

The reader who has worked through Volumes I through XI and completed the capstone has the discipline to engage with Volume XII substantively. The discipline of design (this volume) is the precondition for the discipline of engineering-at-civilization-scale (Volume XII): the engineer who does not exercise discipline on individual projects cannot meaningfully address the cumulative consequences of engineering work. The capstone graduates the reader into the discipline; Volume XII presents the discipline's largest expression.

The book's thesis pinned at the start of Volume I: \emph{Engineering is the discipline by which measured reality becomes reliable intervention under constraint, failure, scale, and responsibility. Engineering is not applied science. Engineering is disciplined intervention.}

The capstone is the reader's first disciplined intervention at the engineering scale this book has been preparing for. Volume XII places the intervention in the context of all the other interventions the engineering profession is responsible for. The reader's engineering career is the lifelong application of the discipline; the capstone is the first substantive piece of it.

\begin{principle}[The capstone is the integration]
The capstone is the integration of every previous volume into the reader's first substantive engineering work. The chapters taught the elements; the studios integrated the elements on bounded problems; the capstone is the reader's integrated practice on a problem of the reader's choice at meaningful scale. The successful completion of the capstone is the reader's confirmation that the book's discipline has become the reader's working practice.
\end{principle}

\begin{archetype}[ethics]
The capstone is the volume's expression of engineering's substantive commitment. The reader's choice of problem reflects the reader's professional priorities; the reader's design choices reflect the reader's integrated discipline; the reader's deliverable reflects the reader's professional standards. The capstone is the reader's first complete commitment as an engineer; the lifetime of subsequent engineering work is the commitment's continuing expression.
\end{archetype}

\begin{project}[The capstone]
Complete the capstone: choose a problem; commit the time; produce the design pack; submit it for review by a panel of qualified persons; reflect on the work and on what it has taught.

\textbf{The problem.} The problem must be real, meaningful, and at scale (Section 2). The reader's commitment to the problem sustains the work over its duration.

\textbf{The time.} The capstone requires substantive time: six months to a year of part-time effort, or several months of concentrated effort. The reader plans the time before beginning; the project's success requires sustained engagement rather than fragmented effort.

\textbf{The design pack.} The deliverable is the complete pack (Section 4). The pack is substantial; its substance is in the engineering work it documents.

\textbf{The review.} The pack is reviewed by a panel of qualified persons (engineering colleagues; professional society representatives; subject-matter experts in the problem's domain). The review's findings inform the pack's revisions and the reader's reflection.

\textbf{The reflection.} The capstone produces, alongside the design pack, the reader's reflection on what the work taught: the technical lessons; the disciplinary lessons; the personal lessons; the professional commitment the work represents.

\textbf{Hazard.} The capstone is the most substantial single project this book asks the reader to undertake. The reader's commitment is the precondition; the discipline is the application; the deliverable is the substantive product. The reader who completes the capstone has done substantive engineering work; the reader who does not has not yet done what this book is preparing the reader to do.
\end{project}
