\chapter{Solutions, electrochemistry, batteries}
\label{vol05:ch03}

\epigraph{The electromotive force of a galvanic cell is determined by
the activities of the ions taking part in the chemical change.}{Walther
Nernst, 1889, \cite[pp.~129--181]{hist:nernst1889}, in the founding
paper of modern electrochemistry.}

\chapmeta{Half-life: foundational for the chemistry; medium-life for commercial battery chemistries (specific energies and costs update every few years). Archetypes: transport (charge), failure. Exercise target: 29. Project track: household battery construction and characterisation. Hazard class: low (galvanic cells at single-digit volts and currents below $100\,\text{mA}$); only the lithium-ion failure discussion involves hazards beyond the project scope. Reader path default: standard, with sections explicitly tagged. Named cases: Boeing 787 lithium-ion battery incidents (2013).}

\noindent
A battery is electrochemistry packaged for engineering use: a spatially
separated redox reaction whose electron transfer goes through an
external circuit and whose ion transport goes through an internal
electrolyte. The chapter develops the thermodynamic and kinetic content
of electrochemical cells, the engineering vocabulary for solutions and
their colligative behaviour, the half-cell potentials that organise the
table of standard electrode potentials, and the commercial battery
chemistries (lead-acid, nickel-cadmium, nickel-metal-hydride, lithium-ion,
lithium iron phosphate, sodium-ion, redox flow) that the working
engineer specifies for electrical storage. The chapter closes with the
failure mode that has dominated battery engineering since the
mass-market entry of high-energy lithium chemistries: thermal runaway,
its mechanism, its propagation, and the containment-and-mitigation
strategies that the post-2013 generation of cell and pack designs has
adopted.

\section{Solutions and concentration}\pathtag{core}

A \engterm{solution} is a homogeneous mixture of two or more components.
The component present in largest quantity is the \engterm{solvent}; the
others are \engterm{solutes}. The engineering use of solutions is
pervasive: aqueous electrolytes carry charge in batteries; aqueous
acids and bases participate in chemical processing (Chapter~2 of this
volume); organic solvents dissolve polymer melts in coating processes;
liquid metals dissolve gases that govern weld quality (Chapter~5 of
this volume).

\engterm{Concentration units}. The chemist and the engineer disagree
on the working unit. The chemist favours \engterm{molarity} $c =
n/V$ (moles of solute per litre of solution) and \engterm{molality} $m
= n/m_{\text{solvent}}$ (moles of solute per kilogram of solvent). The
engineer often uses \engterm{mass fraction} $w_i = m_i/m_{\text{total}}$
(dimensionless) or \engterm{volume fraction} $\phi_i = V_i/
V_{\text{total}}$. The mole fraction $x_i = n_i/n_{\text{total}}$
appears in equilibrium calculations. The four units coincide
in the dilute limit (after appropriate scaling) and diverge at higher
concentrations; a manufacturer's data sheet uses one of them and the
reader who imports the number must verify which.

\engterm{Ideal vs non-ideal solutions}. An \engterm{ideal solution}
obeys Raoult's law: the partial pressure of each component above the
liquid is the product of its mole fraction in the liquid and the
saturation pressure of the pure component, $p_i = x_i p_i^*$.
Real solutions deviate; the ratio $p_i/(x_i p_i^*)$ is the
\engterm{activity coefficient} $\gamma_i$, and the activity
$a_i = \gamma_i x_i$ replaces the mole fraction in non-ideal
equilibrium expressions. The activity coefficient is unity in the
infinitely dilute limit (for the solute) and unity for the pure
solvent at $x = 1$. Solutions with positive deviation from Raoult's
law ($\gamma > 1$) are those in which the solute-solvent attraction is
weaker than the average of solute-solute and solvent-solvent
attractions; negative deviation ($\gamma < 1$) is the opposite. The
engineering content is in distillation (Volume IV Ch~7) and in
solubility calculations.

\engterm{Solubility}. The equilibrium concentration of a solute in a
solvent at a given temperature is the solute's solubility. Solubility
in water is broadly correlated with chemical similarity: polar solutes
dissolve in polar solvents, non-polar solutes in non-polar solvents
(the ``like dissolves like'' rule of thumb). For ionic compounds, the
solubility is set by the lattice energy on one side and the
hydration energy of the constituent ions on the other; the difference
of the two determines whether the compound dissolves and by how much.
Salts of small, highly charged ions (LiF, MgO, Al$_2$O$_3$) have
large lattice energies relative to their hydration energies and are
sparingly soluble; salts of large, singly charged ions (NaCl, KCl,
KNO$_3$) have well-matched lattice and hydration energies and dissolve
readily.

\engterm{Colligative properties}. Four properties of dilute solutions
depend on the number of solute particles rather than on their identity:
boiling-point elevation, freezing-point depression, osmotic pressure,
and vapour-pressure lowering. The colligative properties are the
working tools of (a) antifreeze formulation, (b) intravenous-fluid
preparation, (c) molecular-weight determination of unknown solutes, and
(d) the design of separation membranes.

For dilute solutions of non-electrolytes, the freezing-point depression
is $\Delta T_f = -K_f m$, with $K_f$ the cryoscopic constant of the
solvent ($1.86\,\si{\kelvin}\cdot\text{kg}/\text{mol}$ for water).
Glycol-water mixtures used as automotive antifreeze freeze at
$-37\,\degC$ near $50\,\%$ ethylene glycol by volume. The boiling-point
elevation is $\Delta T_b = K_b m$, with $K_b = 0.512\,\si{\kelvin}\cdot
\text{kg}/\text{mol}$ for water. The osmotic pressure is $\Pi = c R T$
(van't Hoff's law for dilute solutions), with $c$ the molar
concentration of solute particles; at body temperature, blood serum
has an osmotic pressure of approximately $7.7\,\text{atm}$ relative to
pure water. The osmotic pressure of an electrolyte solution is larger
than its formula concentration would predict by the van't Hoff factor
$i$, approximately equal to the number of ions per formula unit
($i \approx 2$ for NaCl, $i \approx 3$ for MgCl$_2$).

\section{Electrochemical cells}\pathtag{core}

An \engterm{electrochemical cell} is a device that converts the free
energy of a redox reaction into electrical work, or that uses electrical
work to drive a non-spontaneous redox reaction. The two configurations
are the \engterm{galvanic cell} (spontaneous, free energy out as
electrical work) and the \engterm{electrolytic cell} (non-spontaneous,
electrical work in to drive a chosen reaction).

A galvanic cell has two electrodes (the \engterm{anode}, where
oxidation occurs, and the \engterm{cathode}, where reduction occurs),
each immersed in its own electrolyte solution, with the two electrolytes
connected by a salt bridge or porous separator that allows ion transport
without bulk mixing. The Daniell cell, the historically important
example, has a zinc electrode in zinc sulphate solution (anode) and a
copper electrode in copper sulphate solution (cathode):
\begin{align*}
\text{Zn} &\to \text{Zn}^{2+} + 2 e^-, \quad &E^\circ = +0.76\,\text{V}, \\
\text{Cu}^{2+} + 2 e^- &\to \text{Cu}, \quad &E^\circ = +0.34\,\text{V}.
\end{align*}
The cell potential is the difference of the cathode and anode standard
potentials, $E^\circ_{\text{cell}} = E^\circ_{\text{cath}} -
E^\circ_{\text{an}} = 0.34 - (-0.76) = 1.10\,\text{V}$ (writing the zinc
half-reaction in the reduction direction to evaluate it consistently).
The sign convention used throughout this chapter follows the IUPAC
convention: standard electrode potentials are reported for the
reduction half-reaction, and the cell potential is the cathode minus
the anode reduction potential.

\engterm{Cell notation}. The IUPAC notation for the Daniell cell is
\[
\text{Zn}|\text{ZnSO}_4(aq, 1\,M)||\text{CuSO}_4(aq, 1\,M)|\text{Cu},
\]
with single vertical bars denoting phase boundaries and the double
bar denoting the salt bridge. The anode is written on the left, the
cathode on the right. The convention preserves the sign of the cell
potential automatically.

\engterm{The Nernst equation}. The actual cell potential at any
composition is the standard potential modified by the reaction
quotient. For a half-reaction with $n$ electrons,
\[
E = E^\circ - \frac{RT}{nF} \ln Q,
\]
where $F = 96\,485\,\text{C}/\text{mol}$ is Faraday's constant
\cite{hist:nernst1889}. At $T = 298\,\si{\kelvin}$, $RT/F = 0.02569\,
\text{V}$ and the equation reduces to
\[
E = E^\circ - \frac{0.0592}{n} \log_{10} Q.
\]
The Nernst equation is the engineer's working tool for predicting how
a cell's voltage changes with concentration, with state of charge in a
battery, and with electrolyte purity.

\engterm{Connecting voltage to free energy}. The maximum electrical
work that can be extracted from a cell is
\[
W_{\text{max}} = -\Delta G = n F E_{\text{cell}}.
\]
The relation is the bridge between chemical thermodynamics
(Volume IV) and electrical engineering (Volume VII): one mole of
electrons carrying one volt of potential difference releases $96.5\,
\text{kJ}$ of work. A cell with $E^\circ = 1.1\,\text{V}$ and a
two-electron transfer per mole of zinc has standard reaction free
energy $-2 \times 96{,}485 \times 1.1 = -212\,\text{kJ}/\text{mol}$.

\engterm{The reversible reference electrode}. A standard hydrogen
electrode (SHE) consists of a platinum electrode immersed in $1\,M$
H$^+$ solution and bathed in $1\,\text{atm}$ H$_2$ gas. The half-reaction
$2\,\text{H}^+ + 2 e^- \to \text{H}_2$ is assigned the reference
potential $E^\circ = 0$ by convention. All other standard electrode
potentials are reported relative to this reference. In practice, the
SHE is awkward; the saturated calomel electrode (Hg/Hg$_2$Cl$_2$ in
saturated KCl, $E = +0.241\,\text{V}$ vs SHE) and the silver/silver
chloride electrode (Ag/AgCl in $3\,M$ KCl, $E = +0.210\,\text{V}$) are
the working laboratory references.

\section{Electrode potentials}\pathtag{standard}

A table of standard electrode potentials organises the redox
chemistry of the elements and the simple ions. The reduction
half-reaction with the more positive $E^\circ$ is the stronger
oxidising agent; the reduction half-reaction with the more negative
$E^\circ$ is the stronger reducing agent (its oxidation form has the
more positive oxidation potential). The most positive standard
reduction potentials in the table (fluorine $F_2 + 2e^- \to 2F^-$,
$E^\circ = +2.87\,\text{V}$; ozone, $+2.07\,\text{V}$; permanganate
in acid, $+1.51\,\text{V}$) are the strongest oxidisers; the most
negative (lithium, $-3.04\,\text{V}$; potassium, $-2.93\,\text{V}$;
calcium, $-2.87\,\text{V}$) are the strongest reducers.

\engterm{Predicting reactions}. A redox reaction is spontaneous when
the cell potential is positive. Zinc metal reduces copper(II) ion
($E^\circ_{\text{cell}} = 1.10\,\text{V}$, spontaneous); copper metal
does not reduce zinc(II) ion (the reverse, $-1.10\,\text{V}$,
non-spontaneous). The table gives a quick prediction for any
combination of reactants whose half-reactions are tabulated. The
prediction is thermodynamic; it does not address rate. A spontaneous
reaction may proceed slowly (the corrosion of aluminium is
thermodynamically favoured but kinetically blocked by a passive
oxide layer) or rapidly (the corrosion of sodium in air is both
favoured and rapid).

\engterm{The galvanic series in seawater}. The ordering of metals by
corrosion potential in seawater is empirically established and is the
working chart for galvanic corrosion in marine engineering. The
positive end (noble) contains gold, platinum, graphite, titanium, and
stainless steels in their passive state; the negative end (active)
contains magnesium, zinc, aluminium, mild steel. When two dissimilar
metals are connected electrically in a common electrolyte, the more
active metal corrodes preferentially (\engterm{galvanic corrosion});
the more noble metal is protected. The engineering uses are
deliberate (zinc sacrificial anodes on ship hulls, magnesium
sacrificial anodes on water heaters) and the engineering hazards are
real (copper plumbing connected to aluminium fittings produces rapid
corrosion of the aluminium). Chapter~9 of this volume develops the
corrosion side; the present chapter establishes the half-cell
machinery.

\engterm{Overpotential}. A real electrode operating at non-zero
current shows a voltage different from its equilibrium Nernstian
value. The difference is the \engterm{overpotential} $\eta = E - E_{eq}$,
positive for a cathode operating at higher current than equilibrium
predicts and the same magnitude but opposite sign for an anode. The
overpotential has three components: \engterm{activation overpotential}
(the kinetic barrier to electron transfer at the electrode surface,
described by the Butler-Volmer equation), \engterm{ohmic overpotential}
(the IR drop across the electrolyte and electrode resistance), and
\engterm{concentration overpotential} (the local depletion of reactant
or accumulation of product at the electrode surface). All three reduce
the deliverable voltage of a battery under load; the engineering of
electrodes and electrolytes is the engineering of minimising
overpotential at the design current density. The \engterm{exchange
current density} $i_0$ is the equilibrium rate of electron transfer
in either direction at zero overpotential; high $i_0$ means low
activation overpotential, and the design of catalysts and electrode
morphologies aims to maximise $i_0$ for the desired half-reaction.

\engterm{Reference electrodes in practice}. Industrial pH meters use
glass electrodes (specific to H$^+$) referenced to internal Ag/AgCl
electrodes. Ion-selective electrodes for sodium, potassium, calcium,
chloride, and fluoride are the working transducers in clinical
chemistry analysers. The Clark electrode for dissolved oxygen and
the Severinghaus electrode for dissolved CO$_2$ are the working
gas-sensing electrochemical devices. The reader who reads off
membrane-electrode-voltage outputs from any of these is implicitly
applying a Nernst-equation-derived calibration.

\section{Electrolysis}\pathtag{standard}

In an \engterm{electrolytic cell}, an external power supply drives a
non-spontaneous reaction. The anode is the positive electrode (driven
to higher potential), the cathode the negative; oxidation occurs at
the anode and reduction at the cathode, the same as in a galvanic cell.
The sign convention reverses for the polarity of the cell relative to
its function: a battery being charged is an electrolytic cell, and the
terminal that was the cathode during discharge becomes the anode during
charge, and vice versa.

\engterm{Faraday's laws of electrolysis}. The mass of a substance
deposited or dissolved at an electrode is proportional to the
quantity of charge passed (Faraday's first law); the masses of
different substances deposited by the same quantity of charge are
proportional to their chemical equivalent weights (Faraday's second
law). Together,
\[
m = \frac{M I t}{n F},
\]
where $m$ is the mass deposited (or evolved), $M$ the molar mass, $I$
the current, $t$ the time, $n$ the electron transfer per formula unit,
and $F$ Faraday's constant. The laws govern industrial electroplating
(copper, nickel, chromium, gold deposition onto conducting substrates),
electrowinning (production of aluminium from molten cryolite-alumina
mixtures, of zinc and copper from aqueous sulphate solutions, of
chlorine from aqueous brine), and electrochemical machining.

\engterm{The chlor-alkali process}. Electrolysis of brine (concentrated
aqueous NaCl) produces chlorine gas at the anode, hydrogen gas at the
cathode, and aqueous sodium hydroxide. The three products are the
high-volume commodity outputs of the chlor-alkali industry, one of the
oldest electrochemical processes in industrial chemistry. Modern
chlor-alkali plants use either diaphragm cells, mercury cells (now
phased out in many jurisdictions because of environmental concerns),
or membrane cells (the modern standard). The cell voltage required
exceeds the thermodynamic minimum ($2.19\,\text{V}$ at standard
conditions) by the overpotentials at each electrode plus the
electrolyte IR drop; total cell voltages of $3$ to $4\,\text{V}$ at
operating current densities of $2$ to $5\,\text{kA}/\text{m}^2$ are
typical.

\engterm{Hall-H\'eroult aluminium}. Aluminium is electrowon from
alumina dissolved in molten cryolite (Na$_3$AlF$_6$) at $\sim 950\,
\degC$. Carbon anodes are consumed as carbon dioxide (each tonne of
aluminium consumes $\sim 0.45\,\text{tonne}$ of carbon), and the
process is highly energy-intensive ($\sim 13$ to $15\,\text{kWh}/
\text{kg}$ of aluminium, compared to a thermodynamic minimum of
$\sim 5.7\,\text{kWh}/\text{kg}$ at the metal-production half-cell
voltage of $1.7\,\text{V}$). Aluminium production is therefore
co-located with cheap electricity (hydroelectric in Norway and
Quebec, geothermal in Iceland, increasingly renewable elsewhere) and
is the largest industrial electrochemical operation in the world.

\engterm{Electroplating}. Coating a metallic or non-metallic substrate
with a thin metal layer by cathodic deposition. The substrate is the
cathode, the plating bath is an aqueous solution of a salt of the
plating metal, and the anode is either a soluble electrode of the
plating metal (which dissolves at the same rate as the deposition) or
an inert electrode (which evolves oxygen instead). Chromium plating
on automotive bumpers, nickel plating on tools, gold plating on
electrical contacts, and zinc plating on steel for corrosion
protection are the dominant industrial applications. The plating
thickness is controlled by the integrated current; uniformity is
controlled by the bath's \engterm{throwing power}, the ability of the
solution to deposit metal in low-current-density regions (geometric
recesses).

\section{Battery chemistries}\pathtag{core}

A battery is a stack of one or more galvanic cells, packaged for
electrical work, sized for an application's voltage and energy
demands. The engineering of a battery is a compromise among specific
energy (Wh/kg), specific power (W/kg), cycle life (number of
charge-discharge cycles to end of life), calendar life, safety
behaviour, cost per kWh, and temperature tolerance. No single
chemistry optimises all six. The dominant commercial chemistries are
covered below; the catalogue is current as of 2025.

\subsection{Lead-acid}

The oldest rechargeable battery, due to Gaston Plant\'e in 1859. The
positive electrode is lead dioxide on a lead grid; the negative
electrode is sponge lead; the electrolyte is sulphuric acid solution
($\sim 5\,M$ in a fully charged cell). The reactions are
\begin{align*}
\text{PbO}_2 + 4\,\text{H}^+ + \text{SO}_4^{2-} + 2 e^- &\to \text{PbSO}_4
+ 2\,\text{H}_2\text{O}, \quad E^\circ_{\text{cath}} = +1.69\,\text{V}, \\
\text{Pb} + \text{SO}_4^{2-} &\to \text{PbSO}_4 + 2 e^-, \quad
E^\circ_{\text{an}} = -0.36\,\text{V},
\end{align*}
giving an open-circuit cell voltage of $2.05\,\text{V}$. Both
electrodes' discharged form is lead sulphate, which is sparingly
soluble; the discharge reaction is a precipitation onto the electrode
surfaces.

Specific energy: $30$ to $50\,\text{Wh}/\text{kg}$ at the cell level.
Cycle life: $200$ to $500$ cycles in deep discharge, $1000$ or more
in shallow discharge service. Temperature range: $-20$ to $+50\,
\degC$. Self-discharge rate: $\sim 5\,\%$ per month. Cost: very low,
$\$100/\text{kWh}$ to $\$200/\text{kWh}$ at the pack level. The
working applications are automotive starting batteries (high-power
short-burst, with thin plates) and stationary backup storage (deep-
cycle service, with thick plates); the design difference is one of
plate geometry, not chemistry. Sealed valve-regulated lead-acid (VRLA)
cells with absorbed-glass-mat (AGM) or gel electrolyte have replaced
flooded cells in many applications because of their lower maintenance
and easier transport. The fundamental limitation is the lead's high
atomic weight; the chemistry is mature and unlikely to improve much
further.

\subsection{Nickel-cadmium}

Nickel-cadmium (NiCd) batteries use a nickel oxyhydroxide cathode, a
cadmium anode, and an alkaline (potassium hydroxide) electrolyte. The
cell voltage is $1.2\,\text{V}$ nominal. Specific energy is $40$ to
$60\,\text{Wh}/\text{kg}$, cycle life $1000$ to $2000$ deep cycles,
temperature range $-40$ to $+60\,\degC$. NiCd cells tolerate abuse
(overcharge, overdischarge, high discharge rate) better than most
other chemistries and were the standard portable battery for power
tools and emergency lighting through the 1990s. The cadmium content
is environmentally hazardous; the Restriction of Hazardous Substances
directive (RoHS) banned new NiCd consumer products in the European
Union from 2006 except for narrow industrial exemptions. NiCd has
been largely replaced by nickel-metal-hydride and lithium-ion in the
consumer and traction markets.

\subsection{Nickel-metal-hydride}

NiMH batteries replace the cadmium anode with a hydrogen-absorbing
metal alloy (typically an AB$_5$ rare-earth-nickel intermetallic such
as LaNi$_5$). Cell voltage is $1.2\,\text{V}$. Specific energy
is $60$ to $90\,\text{Wh}/\text{kg}$. Cycle life is $500$ to $1000$.
Self-discharge of conventional NiMH is high ($\sim 30\,\%$ per month);
low-self-discharge NiMH chemistries developed in the 2000s reduce
this to $\sim 10\,\%$ per year (the ``eneloop'' family is the
canonical commercial example). NiMH was the working chemistry for the
first-generation hybrid vehicles (Toyota Prius from 1997, Honda Civic
Hybrid from 2002) before lithium-ion displaced it in the late 2010s;
NiMH remains common in consumer AA and AAA rechargeables and in some
hand tools.

\subsection{Lithium-ion}

The dominant high-energy rechargeable chemistry since the early 2000s.
A lithium-ion cell shuttles Li$^+$ ions between an intercalation
cathode and an intercalation anode, with the electrons travelling
through the external circuit. The chemistry was made commercially
practical by Goodenough's identification of LiCoO$_2$ as a viable
cathode \cite{hist:goodenough1980} and by Yoshino's identification of
graphite as a viable anode. Sony commercialised the first
LiCoO$_2$-graphite cell in 1991; the 2019 Nobel Prize in Chemistry
recognised Goodenough, Whittingham, and Yoshino for the chemistry.

The lithium-cobalt-oxide cell with graphite anode has nominal cell
voltage $3.6\,\text{V}$, specific energy $150$ to $250\,\text{Wh}/
\text{kg}$ at the cell level, cycle life $500$ to $1000$, and a
temperature range of $0$ to $45\,\degC$ for charging and $-20$ to
$+60\,\degC$ for discharging. The chemistry is the working battery of
laptops, smartphones, and consumer electronics. The reactions are
\begin{align*}
\text{LiCoO}_2 &\rightleftharpoons \text{Li}_{1-x}\text{CoO}_2 + x\,
\text{Li}^+ + x e^- \quad \text{(cathode)}, \\
\text{C}_6 + x\,\text{Li}^+ + x e^- &\rightleftharpoons \text{Li}_x\text{C}_6
\quad \text{(anode)}.
\end{align*}
Variants on the cathode chemistry (nickel-cobalt-aluminium NCA,
nickel-manganese-cobalt NMC in several stoichiometries, the latest
NMC-811 with $80\,\%$ nickel content) trade thermal stability against
specific energy.

\subsection{Lithium iron phosphate}

LiFePO$_4$ (LFP) replaces the layered cobalt oxide cathode with an
olivine-structured iron phosphate. The cell voltage is $3.2$ to
$3.3\,\text{V}$, lower than LiCoO$_2$ but more stable thermally: the
LiFePO$_4$ cathode does not release oxygen on overheating, eliminating
the principal feed-forward mechanism for thermal runaway. Specific
energy is $90$ to $140\,\text{Wh}/\text{kg}$, lower than NCA or NMC.
Cycle life is $2000$ to $5000$ deep cycles, substantially longer than
the layered-oxide chemistries. Cost per kWh at the pack level, as of
2025, has fallen below $\$80/\text{kWh}$ in volume production; the
chemistry has consequently displaced layered-oxide chemistry in
stationary storage and in entry-level electric vehicles. LFP's
combination of safety, cycle life, and cost has made it the dominant
grid-scale lithium chemistry, current as of 2025.

\subsection{Sodium-ion}

Sodium-ion cells use the same intercalation architecture as
lithium-ion, with sodium replacing lithium and the cathode and anode
materials adjusted to accommodate the larger Na$^+$ ion. Commercial
products entered the market around 2023 with specific energies in the
$120$ to $160\,\text{Wh}/\text{kg}$ range, comparable to LFP but with
lower cost (sodium is much cheaper than lithium and is geographically
distributed rather than concentrated in a few producing regions).
Cell voltage is $\sim 3.0\,\text{V}$. The chemistry is positioned for
stationary storage and for low-end vehicles where the cost advantage
matters more than the modest specific-energy gap to lithium-ion. The
production volumes and cost trajectories of sodium-ion are evolving
rapidly as of 2025; any specific number here is medium-life rather
than durable.

\subsection{Redox flow}

A \engterm{flow battery} separates the energy-storing electrolyte from
the cell stack. Two solutions, one for each electrode, are pumped
through a stack of cells with ion-exchange membranes; the energy
storage scales with the electrolyte tank volume, while the power
scales with the cell stack area. The vanadium redox-flow battery
(V$^{2+}$/V$^{3+}$ on one side, VO$_2^+$/VO$^{2+}$ on the other) is
the most commercially developed chemistry; specific energy at the
electrolyte level is $20$ to $40\,\text{Wh}/\text{kg}$, cycle life is
$> 10\,000$ cycles with negligible capacity fade. The independent
scaling of energy and power makes flow batteries attractive for
multi-hour stationary storage; the high system cost (membranes,
pumps, electrolyte) makes them poorly suited for small portable
applications. Other flow chemistries (zinc-bromine, iron-chromium,
organic quinone-based) are under development as of 2025.

\section{Fuel cells}\pathtag{standard}

A \engterm{fuel cell} is a galvanic cell whose reactants are supplied
continuously from external storage rather than from the cell's own
electrode mass. The cell does not store energy; it converts the
chemical energy of the fuel directly to electrical work. The
canonical reaction is hydrogen-oxygen,
\[
2\,\text{H}_2 + \text{O}_2 \to 2\,\text{H}_2\text{O}, \quad
E^\circ_{\text{cell}} = 1.23\,\text{V},
\]
with $\Delta H_r = -286\,\text{kJ}/\text{mol H}_2$ (referenced to
liquid water at the standard state) and standard cell voltage
$1.23\,\text{V}$. The reaction is exothermic; the difference between
$-\Delta H$ and $-\Delta G = nFE^\circ$ is rejected as heat, with the
maximum efficiency (Gibbs to enthalpy) of $83\,\%$ at the standard
state. The actual efficiency at operating current density is lower
because of activation, ohmic, and concentration overpotentials.

\engterm{Proton-exchange-membrane (PEM) fuel cells} use a
sulphonated-fluoropolymer membrane (DuPont's Nafion is the canonical
material) as the proton conductor. Operating temperature is $60$ to
$90\,\degC$, set by the membrane's water-management requirement.
Platinum-on-carbon catalysts are required at both electrodes; the
expense of the platinum loading and the membrane is the dominant cost
of the cell stack. PEM fuel cells are the working chemistry for
hydrogen-powered automotive applications (Toyota Mirai, Hyundai
Nexo) and for portable backup-power applications. The fuel-cell
stack itself has specific power of order $1\,\text{kW}/\text{kg}$;
the full vehicle system, including hydrogen tank, balance of plant,
and battery hybridisation, has specific energy comparable to a
lithium-ion-battery-only system at $\sim 2025$ technology levels.

\engterm{Solid-oxide fuel cells (SOFCs)} use a ceramic
oxide-ion-conducting electrolyte (yttria-stabilised zirconia is the
canonical material) at $700$ to $1000\,\degC$. The high operating
temperature eliminates the need for noble-metal catalysts and allows
direct use of hydrocarbon fuels (the cell can internally reform
methane to hydrogen and carbon monoxide). Stationary SOFCs are
deployed in distributed generation applications; the high operating
temperature is incompatible with portable applications.

\engterm{Alkaline fuel cells} use a concentrated KOH electrolyte and
non-noble-metal catalysts (nickel, silver). They were the original
fuel-cell chemistry, used in the Apollo and Space Shuttle programs.
The chemistry is intolerant of carbon dioxide in the air feed
(carbonate formation poisons the electrolyte), which has limited
terrestrial deployment.

\engterm{Phosphoric-acid fuel cells} use concentrated H$_3$PO$_4$ as
the electrolyte at $\sim 200\,\degC$. The cells are robust and
long-lived ($> 40\,000\,\text{h}$ in field service) but have modest
specific energy. The chemistry was the working choice for the first
generation of stationary commercial fuel-cell installations in the
1990s and is still in service in some baseload-generation
applications.

\section{Failure: thermal runaway in lithium-ion cells}\pathtag{core}

The chapter's failure section names the failure mode that has shaped
the engineering of every modern high-energy battery: thermal runaway
in lithium-ion cells. The mechanism is more specific than the general
runaway-reaction pattern of Chapter~2 of this volume; the engineering
mitigations are correspondingly specific.

\engterm{The mechanism}. A lithium-ion cell stores energy in the
chemical potential difference between the lithiated anode (LiC$_6$)
and the delithiated cathode (Li$_{1-x}$MO$_2$). The flammable
constituents of the cell are the organic carbonate electrolyte
(ethylene carbonate, dimethyl carbonate, diethyl carbonate, ethyl
methyl carbonate, all of which are organic liquids with flash points
below $100\,\degC$) and the lithiated graphite anode (lithium metal
incorporated into the graphite layers). The cathode metal oxide
releases oxygen on thermal decomposition; the released oxygen
oxidises the electrolyte and the lithiated graphite, with the
exothermic combustion of the electrolyte and the carbon supplying the
energy that drives the cell to higher temperatures.

The runaway proceeds in stages, with characteristic temperatures
\cite[pp.~71--75]{acc:ntsb-aar14-01-787-battery}:

\begin{itemize}[leftmargin=*]
\item Above $\sim 80\,\degC$: the solid-electrolyte interphase (SEI)
  layer on the anode decomposes, exposing fresh lithiated graphite.
\item Above $\sim 120\,\degC$: the anode-electrolyte reaction begins
  to release heat at a rate that the cell cannot dissipate through
  ordinary heat-transfer paths.
\item Above $\sim 180\,\degC$: the separator (typically polyethylene
  or polypropylene) melts and short-circuits the cell internally.
\item Above $\sim 200\,\degC$: the cathode releases oxygen, and the
  exothermic cathode-electrolyte-anode reaction enters thermal runaway.
\item Above $\sim 600\,\degC$: the cell vents hot gases, electrolyte
  mist, and, if external air is admitted, flames.
\end{itemize}

\engterm{Initiating events}. The trigger for the cascade can be any
of: internal short circuit (lithium dendrite growth penetrating the
separator, or a manufacturing defect such as a metallic particle
embedded in the electrode), external short circuit, mechanical
penetration (nail, crash damage), overcharge (driving the cathode
beyond its safe lithium-removal limit and plating lithium metal on the
anode), overdischarge (which can deposit copper from the anode current
collector and produce internal shorts on recharge), and external
heating (a neighbouring fire, an electronic-component failure that
deposits heat into the cell pack).

\engterm{Cell-to-cell propagation}. A multi-cell pack can sustain a
runaway in one cell without immediately propagating to neighbouring
cells if the inter-cell thermal coupling is weak enough. If the
coupling is strong (cells in direct mechanical contact, with no
insulating spacers, or cells sharing a heatsink), heat from the
runaway cell raises neighbouring cells above their own runaway
threshold, and the cascade propagates through the pack. The Boston
incident of 7 January 2013 aboard a Japan Airlines Boeing 787 is the
canonical demonstration of cell-to-cell propagation in a transport-
aircraft battery installation: an internal short in cell~5 of an
eight-cell pack drove that cell into runaway, the released heat
propagated to cell~6 next, then to cells~3 and~7
\cite[pp.~98--104]{acc:ntsb-aar14-01-787-battery}. The fire consumed
the pack, escaped the battery enclosure, and damaged the
electronic-equipment bay. The FAA grounded the worldwide 787 fleet for
three months while Boeing redesigned the battery enclosure
\cite[pp.~2, 60]{acc:ntsb-aar14-01-787-battery}.

\engterm{The post-2013 design template}. The investigation of the 787
event identified that the cell-level certification had not adequately
tested for cell-to-cell propagation \cite[pp.~111--120]{acc:ntsb-aar14-01-787-battery}.
The redesigned 787 battery accepts that propagation may occur and
contains the consequences via a steel enclosure vented overboard.
The same containment philosophy has spread to other high-energy
transport-aircraft installations and to electric-vehicle traction
batteries: thermal-runaway propagation is treated as a possibility to
be contained rather than an event to be prevented. The engineering
hierarchy of mitigations becomes (a) inherent safety (LFP cathode
chemistry, lower-energy chemistries), (b) cell-level passive safety
(internal current interrupt devices, vent disks, ceramic separators
that resist thermal short circuit), (c) pack-level passive safety
(thermal-insulating spacers between cells, fire-resistant enclosures,
overboard vent paths), and (d) active safety (battery-management
systems that detect anomalous voltage or temperature and disconnect
the cell or pack). A 2016 series of consumer smartphone product
recalls following several confirmed thermal-runaway events traced to
a manufacturing defect that produced internal shorts in a specific
cell vendor's cells; the response was a worldwide product recall and
the introduction of additional cell-screening checks in subsequent
products. The engineering lesson is general: in a high-energy cell
chemistry, the rate at which a manufacturing-defect-induced internal
short evolves into thermal runaway is faster than any reasonable
service-side intervention; the mitigation must be at the cell, pack,
and enclosure levels, not in field service.

\begin{principle}[Energy density buys runaway risk]
The specific energy of a battery chemistry sets an upper bound on the
adiabatic temperature rise of a failed cell. A high-specific-energy
chemistry (NMC at $250\,\text{Wh}/\text{kg}$) has adiabatic temperature
rise on full self-discharge of $\sim 800\,\si{\kelvin}$ above ambient;
a lower-specific-energy chemistry (LFP at $130\,\text{Wh}/\text{kg}$
combined with the absence of cathode oxygen release) has
correspondingly lower runaway energy and lower runaway temperature.
The engineering trade-off between specific energy and runaway hazard
is not eliminable; it is managed by choosing a chemistry whose runaway
behaviour is compatible with the application's containment, by
designing cell and pack mitigations that contain the worst-case event,
and by sizing pack-level enclosures for the unmitigated propagation
scenario.
\end{principle}

\begin{estimation}
\textbf{Question.} Estimate the open-circuit voltage of a lemon
battery built with a zinc-coated screw as one electrode and a copper
wire as the other.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The half-reactions are
\begin{align*}
\text{Zn} &\to \text{Zn}^{2+} + 2 e^-, \quad E^\circ = -0.76\,\text{V}
\text{ (reduction)}, \\
\text{Cu}^{2+} + 2 e^- &\to \text{Cu}, \quad E^\circ = +0.34\,\text{V}.
\end{align*}
The standard cell potential is $E^\circ_{\text{cell}} = E^\circ_{\text{cath}}
- E^\circ_{\text{an}} = 0.34 - (-0.76) = 1.10\,\text{V}$.

The lemon does not contain $1\,M$ Cu$^{2+}$ or Zn$^{2+}$. The lemon
juice contains citric acid ($\sim 0.3\,M$ total), giving pH $\sim 2.5$.
The actual cathode half-reaction is not the Cu$^{2+}$/Cu reduction
(there are essentially no Cu$^{2+}$ ions in solution) but rather the
hydrogen-evolution reaction $2\,\text{H}^+ + 2 e^- \to \text{H}_2$ at
the copper surface, with $E$ shifted by the Nernst equation to
$E = -0.0592 \times \text{pH} = -0.0592 \times 2.5 = -0.15\,\text{V}$
at the operating pH.

The anode is the same Zn/Zn$^{2+}$ couple. The Nernst correction for
trace Zn$^{2+}$ ($\sim 10^{-3}\,M$ in a fresh lemon) is
$E = -0.76 + (0.0592/2) \log_{10}(10^{-3}) = -0.76 - 0.089 = -0.85\,
\text{V}$ (reduction direction).

The cell potential is $E_{\text{cell}} = -0.15 - (-0.85) = 0.70\,
\text{V}$.

\smallskip
\textit{Postmortem.} A typical measured lemon-battery open-circuit
voltage is $0.85$ to $0.95\,\text{V}$ for a zinc-copper cell. The
estimate is somewhat low. The discrepancy comes from the
overpotential of hydrogen evolution on copper (which is non-negligible
even at near-zero current), from the fact that the actual cathode
reaction may involve dissolved oxygen rather than hydrogen evolution
(the O$_2$/H$_2$O couple has $E^\circ = +1.23\,\text{V}$, much more
positive than H$^+$/H$_2$), and from the simplifications of treating
the lemon as a well-stirred electrolyte. The estimate is within $25\,
\%$ of the measured value, which is the expected accuracy for a
back-of-envelope Nernst calculation. The result is also a useful
upper-bound check: a lemon battery cannot produce a voltage exceeding
the standard $1.1\,\text{V}$ of the Daniell cell, and the actual
voltage is reduced by the deviations from $1\,M$ ion activity.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Build a lemon battery; characterise its capacity]
The reader constructs a fruit-electrolyte galvanic cell with
household materials, measures its open-circuit voltage and short-
circuit current, and estimates its energy capacity by powering a
low-power LED for a measured duration.

\textbf{Track}. Household electrochemistry.
\textbf{Hazard class}: low. No reactants beyond fruit, common metals,
and a low-voltage LED.

\textbf{Procedure}.

\begin{enumerate}[leftmargin=*]
\item Construct a single cell: insert a galvanised (zinc-coated)
  steel nail or screw and a copper wire or coin into the same lemon,
  separated by a few centimetres. Connect a multimeter across the two
  electrodes and read the open-circuit voltage and the short-circuit
  current.
\item Construct a cell with magnesium (a sharpener blade, if
  conveniently available, or a magnesium-alloy bicycle part) replacing
  the zinc. Read the open-circuit voltage and compare to the zinc cell.
  Predict the value from the standard electrode potentials.
\item Connect four cells in series with copper wires. Verify that the
  series voltage equals the sum of the single-cell voltages within
  the multimeter's accuracy.
\item Power a $1.7\,\text{V}$ red LED from the four-cell stack. The
  LED's typical forward current is $5$ to $20\,\text{mA}$; verify
  whether the cell stack can supply this. If not, identify the
  limiting cell impedance.
\item Measure the cell's discharge characteristic: replace the LED
  with a known resistor ($100\,\Omega$ to $1\,\text{k}\Omega$) and
  log the voltage across the resistor every $5\,\text{min}$ for at
  least one hour. Plot voltage vs time and compute the cumulative
  energy delivered.
\item Estimate the cell's specific energy in Wh/kg from the
  cumulative energy and the cell's total mass (lemon plus electrodes).
  Compare to commercial chemistries.
\end{enumerate}

\textbf{Deliverable}. A five- to eight-page report with the data
table, the discharge curve, a comparison of measured to predicted
voltage, and a brief discussion of which factors (concentration
overpotential, electrode kinetics, internal resistance) dominated the
voltage drop under load.

\textbf{Time}. Two to three hours of bench work, two hours of
write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A $1.0\,M$ aqueous solution of NaCl is prepared by dissolving
$58.5\,\text{g}$ of NaCl in enough water to make $1.00\,\text{L}$ of
solution. Compute the molality (assuming a final solution density of
$1.040\,\text{g}/\text{mL}$) and the mass fraction.
\end{exercise}

\begin{exercise}
Compute the freezing-point depression of a $30\,\%$-by-mass aqueous
ethylene glycol solution. Use $K_f = 1.86\,\si{\kelvin}\cdot\text{kg}/
\text{mol}$ and ethylene glycol molar mass $62.07\,\text{g}/\text{mol}$.
\end{exercise}

\begin{exercise}
For the cell $\text{Zn} | \text{Zn}^{2+}(0.10\,M) ||
\text{Cu}^{2+}(0.50\,M) | \text{Cu}$, compute the cell voltage at
$25\,\degC$ using the Nernst equation.
\end{exercise}

\begin{exercise}
The specific energy of a lithium-ion cell with LiCoO$_2$ cathode and
graphite anode is $180\,\text{Wh}/\text{kg}$ at the cell level. Compute
the equivalent energy in mass of TNT ($4.2\,\text{MJ}/\text{kg}$) and
in litres of gasoline ($35\,\text{MJ}/\text{L}$).
\end{exercise}

\begin{exercise}
A lead-acid battery has nominal voltage $12\,\text{V}$, capacity
$50\,\text{Ah}$, and mass $15\,\text{kg}$. Compute its specific energy
in Wh/kg. Compare to a lithium-ion pack of the same nominal voltage
and capacity at $180\,\text{Wh}/\text{kg}$ at the cell level (assume
$70\,\%$ system-level energy density).
\end{exercise}

\begin{exercise}
Compute the mass of aluminium that can be produced in one hour by an
electrolytic cell drawing $200\,\text{kA}$ at $4.5\,\text{V}$. Use
Faraday's law and the three-electron reduction
$\text{Al}^{3+} + 3 e^- \to \text{Al}$, with molar mass $26.98\,
\text{g}/\text{mol}$. Compute the electrical energy required per
kilogram of aluminium.
\end{exercise}

\begin{exercise}
A PEM fuel cell operates at $0.75\,\text{V}$ and draws $0.5\,\text{A}/
\text{cm}^2$. Compute the cell's voltage efficiency relative to the
standard $1.23\,\text{V}$ and the power density per electrode area.
\end{exercise}

\begin{exercise}
The osmotic pressure of human blood serum at $37\,\degC$ is $7.7\,
\text{atm}$. Compute the equivalent molar concentration of dissolved
particles. State why intravenous fluids are formulated at this
osmotic strength.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the Nernst equation $E = E^\circ - (RT/nF) \ln Q$ from the
relation $\Delta G = -nFE$ and the non-standard free-energy expression
$\Delta G = \Delta G^\circ + RT \ln Q$.
\end{exercise}

\begin{exercise}
Derive Faraday's law of electrolysis $m = MIt/(nF)$ from the definition
of current and the stoichiometry of the electrode reaction.
\end{exercise}

\begin{exercise}
Derive the relation between cell voltage and Gibbs free energy
$\Delta G = -nFE$ by computing the maximum work obtainable from a
cell that moves one mole of $n$-charged species across a potential
difference $E$.
\end{exercise}

\begin{exercise}
Show that the open-circuit voltage of a concentration cell (same
metal on both sides, different ion concentrations) is $E = (RT/nF)
\ln(c_{\text{high}}/c_{\text{low}})$.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the total energy stored in a fully charged Tesla Model 3
Long Range battery pack ($75\,\text{kWh}$ nominal capacity).
Compare to the chemical energy in $1\,\text{L}$ of gasoline and to
the kinetic energy of the vehicle at highway speed.
\end{exercise}

\begin{exercise}
Estimate the heat released when a $50\,\text{kWh}$ lithium-ion
electric-vehicle battery undergoes complete thermal runaway. Compare
to the heat released by combustion of $1\,\text{kg}$ of gasoline
($46\,\text{MJ}$).
\end{exercise}

\begin{exercise}
Estimate the maximum theoretical specific energy of a
hydrogen-oxygen fuel cell using the heat of combustion of hydrogen
and the mass of water produced per mole of hydrogen.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a simple equivalent-circuit model of a lithium-ion cell
(ideal voltage source in series with an ohmic resistance and a
parallel RC). Fit the model parameters to a measured discharge curve.
Identify the regimes in which the simple model fails.
\end{exercise}

\begin{exercise}
Implement a Butler-Volmer model of an electrode at variable
overpotential and current density. Plot the polarisation curve and
identify the activation, ohmic, and concentration regions.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify a battery for a $5\,\text{kW}$ residential photovoltaic
system with daily energy storage of $15\,\text{kWh}$ and ten-year
calendar-life target. Compare LFP, lead-acid, and vanadium-flow
chemistries on cost per kWh, cycle life, and footprint.
\end{exercise}

\begin{exercise}
Specify the battery pack for a $100\,\text{kg}$ delivery drone with
a $30\,\text{min}$ flight time at $10\,\text{kW}$ peak power.
Identify the cell chemistry, the cell count, and the pack-level
thermal-management requirements.
\end{exercise}

\begin{exercise}
Specify a sacrificial anode system for the corrosion protection of
a buried steel pipeline with surface area $1000\,\text{m}^2$ in a
soil of resistivity $5000\,\Omega\cdot\text{cm}$. Identify the
anode metal, the number of anodes, and the expected service life.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A lithium-ion cell's nominal capacity is $2.5\,\text{Ah}$ at C/5
discharge. The measured capacity at $5\text{C}$ discharge is $1.6\,
\text{Ah}$. Identify the dominant loss mechanism and estimate the
cell's internal resistance.
\end{exercise}

\begin{exercise}
A vendor advertises a battery with energy density of $400\,\text{Wh}/
\text{kg}$ at $\$50/\text{kWh}$ in mass production. Identify which
specifications would falsify the claim if measured and propose a
single experimental check.
\end{exercise}

\begin{exercise}
A consumer power tool's battery pack delivers full power at room
temperature and fails to start at $-10\,\degC$. Identify the
mechanism (lithium-ion charging is limited at low temperature
because plating of metallic lithium becomes thermodynamically
preferred over intercalation) and propose a mitigation.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A lithium-ion battery pack in an electric scooter catches fire while
the scooter is parked at a charging station, with no occupant nearby.
Identify the most likely initiating event class (overcharge, internal
short, external short, mechanical damage, external heat) and the
single most useful diagnostic data point that the battery-management
system should have logged.
\end{exercise}

\begin{exercise}
A grid-scale battery installation experiences a fire in one
container. The neighbouring containers, separated by $5\,\text{m}$
of open space, do not ignite. Identify the design feature that
prevented propagation and the failure mode that would have made
propagation likely.
\end{exercise}

\begin{exercise}
A consumer device using a single $18650$ lithium-ion cell experiences
a thermal-runaway event during normal operation. The cell was
manufactured by a vendor with documented quality-control issues.
Identify the regulatory and engineering response chain that the
device manufacturer should follow.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A colleague proposes that the principal safety issue with lithium-ion
batteries is the lithium metal itself. Identify the correction (the
working chemistries have lithium present only as intercalated ion,
not as metal, except in lithium-plating failure modes) and the
correct identification of the dominant flammable constituent.
\end{exercise}

\begin{exercise}
A vendor claims a sodium-ion battery has higher specific energy than
a lithium-ion battery of equivalent cost. Identify the thermodynamic
constraint on the sodium-ion specific energy and the experimental
test that would settle the comparison.
\end{exercise}

\begin{exercise}
A reader argues that fuel cells are ``always more efficient'' than
combustion engines. Identify the qualification this claim requires
(direct chemical-to-electrical conversion versus thermal-engine
Carnot bound) and the condition under which the claim is correct.
\end{exercise}
