\chapter{Electromagnetism: Maxwell's equations as four observations}
\label{vol04:ch08}

\epigraph{The agreement of the results seems to show that light and
magnetism are affections of the same substance, and that light is an
electromagnetic disturbance propagated through the field according to
electromagnetic laws.}{James Clerk Maxwell, \emph{A Treatise on
Electricity and Magnetism}, 3rd ed.\ (Oxford: Clarendon, 1892),
vol.~II, art.~781 \cite[vol.~II, p.~431]{hist:maxwell1873}.}

\chapmeta{Half-life: durable. Archetypes: balance, transport, interface. Exercise target: 31. Project track: physical measurement plus analysis (hybrid). Hazard class: low (bench-scale currents and voltages, mains avoided). Reader path default: standard, with sections explicitly tagged. Named cases: lightning-induced surge damage to grid and electronic equipment (described by mechanism, no single accident named).}

\noindent
The four Maxwell equations are observations of the world before they
are mathematics. Each was discovered as a regularity in experiment, and
each retains the working content of its experimental origin. We state
the four equations in their integral form, derive their differential
form, show that they imply the wave equation, and end with the
failure modes of every electrical system: ground loops, induced
electromagnetic interference, and lightning. The reader who finishes
the chapter writes energy balances across electromagnetic fields with
the same discipline applied in earlier volumes to mass and momentum.

\section{Charge, current, and fields}\pathtag{core}

Electric charge is a property of matter measured in coulombs, $1\,\si{
\coulomb} = 1\,\si{\ampere\second}$. The charge of a proton is
$+e = 1.602\,176\,634\times 10^{-19}\,\si{\coulomb}$, fixed exactly in
the 2019 SI redefinition; the charge of an electron is $-e$. Macroscopic
objects carry net charge equal to the algebraic sum of their elementary
charges. Charge is conserved: the total charge inside any closed
surface changes only by the flux of charge through the surface.

\engterm{Current} is the rate at which charge crosses an oriented
surface,
\[
I = \dv{Q}{t}.
\]
For a wire carrying $1\,\si{\ampere}$, about $6.24\times 10^{18}$
elementary charges cross any cross-section per second. The drift speed
of conduction electrons in a typical copper wire is on the order of
$10^{-4}\,\si{\meter\per\second}$; the electrons are slow, but their
density is high.

The \engterm{current density} $\vect{J}$ is the local current per unit
area, with direction along the flow,
\[
I = \int_{S} \vect{J}\cdot \dd\vect{A},
\]
so $\vect{J}$ carries units $\si{\ampere\per\meter\squared}$. Charge
conservation, applied to an infinitesimal control volume, gives the
\engterm{continuity equation}:
\[
\pdv{\rho}{t} + \divg \vect{J} = 0,
\]
where $\rho$ is the charge density in $\si{\coulomb\per\meter\cubed}$.
The continuity equation is the same accounting equation that closed the
mass balance in Volume IV, Chapter 7: stuff in minus stuff out equals
change in stuff stored, with $\vect{J}$ playing the role of mass flux
and $\rho$ the role of mass density. Every Maxwell equation can be read
as a balance once the right field is identified as flux and the right
field as source.

The relation $I = n q v_d (w t)$ that ties current to the drift of carriers
carries a lesson worth stating in full, because it separates three speeds
that are routinely confused. The current density is $\vect{J} = n q\vect{v}_d$
with $n$ the carrier number density, $q$ the charge per carrier, and
$\vect{v}_d$ the average drift velocity. For copper, $n \approx 8.5\times
10^{28}\,\si{\per\meter\cubed}$, so carrying $1\,\si{\ampere}$ through a
$1\,\si{\milli\meter\squared}$ wire needs a drift of only $v_d = J/(nq) =
(10^{6})/[(8.5\times 10^{28})(1.6\times 10^{-19})] \approx 7\times 10^{-5}\,
\si{\meter\per\second}$, tens of micrometres a second, slower than a
minute hand. The individual electrons, in contrast, move at the Fermi speed
of order $10^{6}\,\si{\meter\per\second}$, but in random directions that
average to the tiny drift. The signal that switches a lamp travels at neither
speed: it is the electromagnetic disturbance in the fields around the wire,
propagating at a fair fraction of the speed of light, so the lamp lights the
instant the switch closes even though the electron that was at the switch
takes an hour to reach it. The three numbers differ by ten orders of
magnitude, and the Poynting result of a later section names which one carries
the energy: neither the crawling drift nor the darting Fermi motion, but the
field in the space around the conductor.

A \engterm{field} is a quantity defined at every point of space. The
electric field $\vect{E}(\vect{r}, t)$ is the force per unit charge that
a test charge would experience at $\vect{r}$ at time $t$,
\[
\vect{F} = q\vect{E},
\]
in units of $\si{\volt\per\meter}$ (equivalently $\si{\newton\per
\coulomb}$). The magnetic field $\vect{B}(\vect{r}, t)$ is the
proportionality between velocity and the velocity-dependent component
of the force on a charge,
\[
\vect{F} = q\vect{v}\times \vect{B},
\]
in units of teslas, $\si{\tesla} = \si{\volt\second\per\meter\squared}$.
The Earth's magnetic field at the surface is around $30$ to $60\,\si{
\micro\tesla}$; a refrigerator magnet is around $5\,\si{\milli\tesla}$;
a clinical MRI machine is $1.5$ to $7\,\si{\tesla}$. Both fields are
observable through the forces they exert; both are real, in the sense
that they store energy that can do work.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-point-charge-field.tex}

The combined force is the \engterm{Lorentz force},
\[
\vect{F} = q(\vect{E} + \vect{v}\times \vect{B}),
\]
the empirical input that closes the loop between fields and motion.
Once we know the fields, the Lorentz force tells us what charges do;
once we know what the charges do, Maxwell's equations tell us what the
fields do. The two halves close.

\begin{archetype}[Balance, transport, interface]
Every electromagnetic equation in the chapter is one of three patterns.
A \engterm{balance} equates a flux across a surface to the source
enclosed (Gauss's law, the no-monopole law). A \engterm{transport}
equation relates the time rate of change of one field to the curl of
another (Faraday, Amp\`ere-Maxwell). An \engterm{interface} condition
relates the field on one side of a material boundary to the field on
the other (boundary conditions on $\vect{E}$, $\vect{D}$, $\vect{B}$,
$\vect{H}$). The discipline of the chapter is to recognise which
pattern a problem requires before reaching for the formula.
\end{archetype}

\section{Gauss's law for electricity}\pathtag{core}

The first Maxwell equation, in integral form, is
\[
\oint_{\partial V} \vect{E}\cdot \dd\vect{A} = \frac{Q_{\text{enc}}}{
\varepsilon_{0}},
\]
where $\partial V$ is a closed surface bounding the volume $V$,
$Q_{\text{enc}} = \int_{V}\rho\,\dd V$ is the charge enclosed, and
$\varepsilon_{0} = 8.854\,187\,\times 10^{-12}\,\si{\farad\per\meter}$
is the permittivity of free space. The law is the experimental content
of Coulomb's $1/r^{2}$ result, extended from point charges to arbitrary
distributions: the total outward flux of $\vect{E}$ through a closed
surface depends only on the charge enclosed.

The differential form follows from the divergence theorem of
Volume II, Chapter 8:
\[
\divg \vect{E} = \frac{\rho}{\varepsilon_{0}}.
\]
The electric field's divergence at a point equals the charge density at
that point, scaled by $1/\varepsilon_{0}$. In a region with no charge
the divergence is zero; field lines neither begin nor end. In a region
with positive charge density, field lines diverge from the charge; with
negative charge density, field lines converge.

\begin{principle}[Choose the symmetry first]
Gauss's law gives the field by quadrature only when the symmetry of the
problem makes $\vect{E}\cdot\dd\vect{A}$ constant on a chosen Gaussian
surface. Three symmetries close the integral:
\begin{itemize}[leftmargin=*]
  \item Spherical (point charge, charged sphere): Gaussian sphere
    centred on the source.
  \item Cylindrical (long wire, charged cylinder): Gaussian cylinder
    coaxial with the source.
  \item Planar (infinite sheet, parallel-plate capacitor):
    Gaussian pillbox spanning the sheet.
\end{itemize}
If none of the three applies, Gauss's law is still true but no longer
a tool for computing the field; the field is obtained instead by
integrating Coulomb's law or by solving Poisson's equation. The
engineer's instinct is to identify the symmetry before reaching for
the integral.
\end{principle}

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-gaussian-surfaces.tex}

A point charge $q$ at the origin produces, by spherical symmetry on
a Gaussian sphere of radius $r$,
\[
4\pi r^{2}\, E(r) = \frac{q}{\varepsilon_{0}}, \quad E(r) =
\frac{q}{4\pi \varepsilon_{0} r^{2}},
\]
which is Coulomb's law recovered. The constant $1/(4\pi\varepsilon_{0})
\approx 8.988\times 10^{9}\,\si{\newton\meter\squared\per\coulomb
\squared}$ is large; even small charges produce large forces at small
separations, which is why static-electricity sparks are uncomfortable
and why most matter is electrically neutral at the macroscopic scale.

An infinite line charge with linear density $\lambda$ in $\si{\coulomb
\per\meter}$ produces, by cylindrical symmetry on a Gaussian cylinder
of radius $r$ and length $L$,
\[
2\pi r L\, E(r) = \frac{\lambda L}{\varepsilon_{0}}, \quad E(r) =
\frac{\lambda}{2\pi \varepsilon_{0} r},
\]
falling as $1/r$ rather than $1/r^{2}$. Inside a coaxial cable the
field between conductors has this form, and the result drives the
cable's capacitance per unit length, derived in Section 8.8.

An infinite sheet with surface charge density $\sigma$ in $\si{\coulomb
\per\meter\squared}$ produces, by planar symmetry on a pillbox
straddling the sheet,
\[
E = \frac{\sigma}{2\varepsilon_{0}},
\]
on each side, normal to the sheet, independent of distance. A pair of
infinite parallel sheets with equal and opposite charge densities
$\pm\sigma$ produces $E = \sigma/\varepsilon_{0}$ between them and zero
outside; this is the working model of the ideal parallel-plate
capacitor.

Gauss's law also settles a result the shielding sections lean on: the field
inside an empty cavity in a conductor is zero, whatever the charges outside.
Take a conductor in static equilibrium, where the field inside the metal must
vanish, because any interior field would drive its free charge until the field
it built cancelled the original. Draw a Gaussian surface entirely within the
metal, wrapping a hollow cavity. The flux through it is zero, since $\vect{E}$
is zero everywhere on it, so the enclosed charge is zero: no net charge sits
on the cavity wall. A sharper argument rules out even a field that averages to
no net charge. Suppose a field line ran across the empty cavity from one wall
to another; its line integral $\int\vect{E}\cdot\dd\vect{\ell}$ along that
line would be nonzero, yet the two endpoints lie in the same conductor at one
common potential, so the integral must be zero, a contradiction unless the
field itself is zero throughout the cavity. The empty cavity is therefore
field-free no matter how strong or how time-varying the external field, up to
the quasi-static limit where the wall's finite conductivity and the cavity's
size against a wavelength begin to matter. This is the electrostatic
\engterm{Faraday cage}: a closed conducting shell holds its interior at a
single potential and screens it from external fields, the reason sensitive
electronics sit in metal boxes and the reason a car shell protects its
occupants from a lightning strike. The magnetic case is harder, because
$\divg\vect{B} = 0$ offers no charge to rearrange and cancel a static field,
which is why the magnetic shielding of a later section must divert flux through
a high-permeability wall rather than cancel it with induced surface charge.
The electric cage is free; the magnetic shield is bought.

\section{No magnetic monopoles}\pathtag{core}

The second Maxwell equation states that the magnetic flux through any
closed surface is zero,
\[
\oint_{\partial V} \vect{B}\cdot \dd\vect{A} = 0,
\]
or equivalently
\[
\divg \vect{B} = 0.
\]
The empirical content is that no magnetic monopole has ever been
observed despite repeated searches in the 20th and 21st centuries; the
upper bound on cosmic monopole flux is so small that for engineering
purposes monopoles do not exist. Every magnetic field line that enters
a closed surface leaves it; magnetic field lines form closed loops.

A consequence is that the magnetic field can be written as the curl of
a vector potential,
\[
\vect{B} = \curl \vect{A},
\]
since the divergence of any curl is zero. The vector potential
$\vect{A}$ is not unique: adding the gradient of any scalar function
leaves $\vect{B}$ unchanged, a freedom called \engterm{gauge
invariance}. The freedom is fixed by a gauge condition; the Coulomb
gauge $\divg \vect{A} = 0$ and the Lorenz gauge $\divg \vect{A} +
\mu_{0}\varepsilon_{0}\,\partial \phi/\partial t = 0$ are the two
most useful.

The no-monopole equation does not directly give a field by quadrature
the way Gauss's law does, but it constrains the topology of every
magnetic field the engineer will encounter. A solenoid's field exits
one end and re-enters the other in a closed-loop pattern; a bar
magnet's field lines emerge from the north pole and return to the
south through space outside and through the magnet inside; a current
loop's field circulates around the loop's axis. The closed-loop
geometry is not optional; it is the consequence of the empirical
absence of monopoles.

The constraint earns its keep in design. A magnetic circuit conserves
flux around every closed path the way a fluid circuit conserves volume
flow: what enters a junction leaves it. This is the magnetic analogue of
Kirchhoff's current law, and it is $\divg\vect{B} = 0$ integrated over a
small volume around the junction. An engineer routing flux through a
branched core sums the fluxes into each node to zero, exactly as a
circuit engineer sums currents, and the reluctance network of the
chapter's later case study rests on this conservation. The same rule
forbids a design from dumping flux into free space at one point and
recovering none of it: every line that leaves the core must return, and
the leakage flux that escapes the intended path is the designer's
constant adversary in transformers and motors, the flux that links the
wrong winding or none at all.

The closed-loop constraint also bends field lines sharply at the surface of
a high-permeability material, the behaviour that lets an iron pole piece
gather a field and a mu-metal shell divert one. At a boundary between two
materials the normal component of $\vect{B}$ is continuous, from
$\divg\vect{B} = 0$ integrated over a thin pillbox, while the tangential
component of $\vect{H}$ is continuous where no free surface current flows,
from Amp\`ere's law on a thin loop. Writing the field angle from the normal
as $\theta$ in each medium, continuity of $B_n = B\cos\theta$ and of $H_t =
(B/\mu)\sin\theta$ gives the refraction law
\[
\frac{\tan\theta_2}{\tan\theta_1} = \frac{\mu_2}{\mu_1},
\]
so a field line crossing from air ($\mu_1 = \mu_0$) into iron ($\mu_2 =
5000\mu_0$) at even a grazing angle emerges almost parallel to the surface
inside the iron, $\tan\theta_2 = 5000\tan\theta_1$, and a line leaving the
iron into air bends sharply toward the normal. The iron therefore behaves
as a flux funnel: field lines in the surrounding air turn to enter it nearly
perpendicular and run along its length inside, which is why a pole piece
concentrates flux into its tip and why the field just outside a
high-permeability surface stands nearly normal to it, the magnetic twin of
the way light bends at a glass surface and the working reason a magnetic
shield must wrap the cavity rather than merely stand beside it.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-solenoid-field.tex}

\begin{mastery}
In a region carrying no free current the field $\vect{H}$ obeys
$\curl\vect{H} = \vect{0}$, so it is the gradient of a scalar, and the
whole apparatus of electrostatics can be borrowed for magnetostatics. Write
$\vect{H} = -\grad\psi_m$ with the \engterm{magnetic scalar potential}
$\psi_m$; then $\divg\vect{B} = 0$ with $\vect{B} = \mu\vect{H}$ in a
uniform region gives Laplace's equation $\laplacian\psi_m = 0$, identical
in form to the electrostatic potential problem, so the same separation of
variables, the same boundary-value solutions, and the same numerical
relaxation solve both. The device is what lets a designer treat a permanent
magnet with a pole-charge model, assigning a fictitious surface magnetic
charge density $\sigma_m = \vect{M}\cdot\unitvec{n}$ to the magnet's faces
and computing the field as if from those charges by Coulomb's law. The
scalar potential carries one hard restriction that the vector potential
does not: it is legitimate only where $\vect{J} = \vect{0}$, because a
region with current has $\curl\vect{H}\neq\vect{0}$ and no single-valued
scalar can represent a field whose line integral around a current-enclosing
loop is nonzero. Trace a loop around a current-carrying wire and $\psi_m$
changes by $\oint\vect{H}\cdot\dd\vect{\ell} = I$ on returning to the start,
so $\psi_m$ is multivalued and useless there. The rule of thumb is that
$\psi_m$ solves the flux paths in the iron and the air gaps of a magnetic
circuit, where the current lives in windings treated as sources at the
boundary, while $\vect{A}$ is the tool the moment current fills the region
of interest, as in a solid conductor or a plasma.
\end{mastery}

\section{Faraday's law}\pathtag{core}

The third Maxwell equation is Faraday's law of induction:
\[
\oint_{C} \vect{E}\cdot \dd\vect{\ell} = -\dv{}{t}\int_{S} \vect{B}
\cdot \dd\vect{A},
\]
where $C$ is a closed curve bounding the surface $S$, and the time
derivative acts on the total magnetic flux through $S$. The line
integral on the left is the \engterm{electromotive force} (EMF)
around $C$; the surface integral on the right is the magnetic flux.
The differential form,
\[
\curl \vect{E} = -\pdv{\vect{B}}{t},
\]
follows from Stokes's theorem. A changing magnetic field produces a
circulating electric field; the minus sign (Lenz's law) is the
empirical content that the induced EMF opposes the change.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-faraday-loop.tex}

Faraday's law is the working principle of every generator, transformer,
and inductive coupling in engineering. The flux through a coil of $N$
turns is $N\Phi_{B}$ where $\Phi_{B} = \int \vect{B}\cdot \dd\vect{A}$
through a single turn; the induced EMF is
\[
\mathcal{E} = -N\dv{\Phi_{B}}{t}.
\]
A transformer with primary voltage $V_{1}$ and turns $N_{1}$ produces
on a secondary with $N_{2}$ turns
\[
V_{2} = V_{1}\frac{N_{2}}{N_{1}},
\]
in the ideal limit of zero leakage flux and zero winding resistance.
The transformer ratio is the source-text reason the grid uses
high-voltage transmission: stepping voltage up at the generator and
down at the consumer reduces $I^{2}R$ losses by the factor $(N_{1}/
N_{2})^{2}$.

The minus sign carries operational consequences. A coil moved into a
field experiences an EMF that drives current in the direction that
\emph{opposes} the motion; the coil feels a retarding force. This is
the principle of regenerative braking in electric vehicles: kinetic
energy is converted to electrical energy by the back-EMF of the
motor's coils, with the mechanical force decelerating the vehicle.
Without the minus sign, kinetic energy could grow without bound from
nothing; the minus sign is conservation of energy in disguise.

\begin{estimation}
\textbf{Question.} A bicycle dynamo lights a $3\,\si{\watt}$ headlamp
when the wheel turns at $20\,\si{\kilo\meter\per\hour}$. Estimate the
magnetic flux change per revolution of the dynamo's rotor and the
peak voltage induced.

\textbf{Estimate, before reading on.} The reader is asked to state a
factor-of-three-confidence answer before reading the working.

\textbf{Working.} The dynamo rotates against the tyre; at $20\,\si{
\kilo\meter\per\hour}$ the wheel's circumference of about $2\,\si{
\meter}$ rotates roughly $3\,\si{\hertz}$. A typical dynamo has a
gear ratio of around $5$ to $10$ between tyre and rotor, giving a
rotor frequency of $15$ to $30\,\si{\hertz}$. The lamp is $3\,\si{
\watt}$ at $6\,\si{\volt}$, so RMS current is $0.5\,\si{\ampere}$ and
peak voltage in a single sinusoid would be about $8.5\,\si{\volt}$.
The peak voltage occurs at maximum $\dd\Phi/\dd t$; with $N$ turns
and peak flux $\Phi_{0}$ in a rotor turning at $\omega$, peak EMF is
$N\Phi_{0}\omega$. For $\omega \approx 2\pi\times 20 = 125\,\si{\per
\second}$ and $V_{\text{peak}} \approx 8.5\,\si{\volt}$, the product
$N\Phi_{0} \approx 0.07\,\si{\weber}$. A coil of $200$ turns then needs
$\Phi_{0} \approx 350\,\si{\micro\weber}$, consistent with a $1\,\si{
\centi\meter\squared}$ pole face seeing $\approx 0.35\,\si{\tesla}$
from a small ferrite magnet.

\textbf{Postmortem.} The estimate hangs on the gear ratio and the
number of turns; both can be observed on a real dynamo. A reader who
guessed within a factor of three internalised the order of magnitude
of generator design without writing a single equation in coordinate
form. The empirical numbers are dyamos sold for bicycles, current as
of 2025.
\end{estimation}

\section{Amp\`ere-Maxwell law}\pathtag{core}

The fourth Maxwell equation, in the form Maxwell himself completed it,
is
\[
\oint_{C} \vect{B}\cdot \dd\vect{\ell} = \mu_{0}\left( I_{\text{enc}}
+ \varepsilon_{0}\dv{}{t}\int_{S}\vect{E}\cdot \dd\vect{A}\right),
\]
or equivalently in differential form
\[
\curl \vect{B} = \mu_{0}\vect{J} + \mu_{0}\varepsilon_{0}\pdv{\vect{E}
}{t},
\]
where $\mu_{0} = 4\pi\times 10^{-7}\,\si{\henry\per\meter}$ is the
permeability of free space (its value redefined by the 2019 SI as
an experimentally determined quantity, no longer exact). The first
term on the right is the current density; the second term is Maxwell's
\engterm{displacement current}, the time rate of change of the
electric field.

The displacement current is the term Maxwell added in 1865 to make the
four equations consistent with charge conservation. Without it,
$\divg \curl \vect{B} = \mu_{0}\divg \vect{J} = -\mu_{0}\partial \rho/
\partial t$, which is non-zero whenever charge accumulates anywhere;
but $\divg \curl$ is identically zero for any field, so the equation
is inconsistent without the displacement current. With the
displacement current,
\[
\divg(\mu_{0}\vect{J} + \mu_{0}\varepsilon_{0}\pdv{\vect{E}}{t}) =
\mu_{0}(-\pdv{\rho}{t}) + \mu_{0}\varepsilon_{0}\pdv{(\rho/
\varepsilon_{0})}{t} = 0,
\]
identically, by continuity. The mathematical requirement of consistency
forced the physical prediction that a changing electric field produces
a magnetic field even in the absence of current, and the prediction
was confirmed empirically by Hertz in 1888 \cite{hist:hertz1893}.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-displacement-current.tex}

For a long straight wire carrying steady current $I$, Amp\`ere's law on
a circle of radius $r$ centred on the wire gives, by cylindrical
symmetry,
\[
2\pi r B(r) = \mu_{0} I, \quad B(r) = \frac{\mu_{0} I}{2\pi r}.
\]
A wire carrying $1\,\si{\ampere}$ at $1\,\si{\centi\meter}$ from its
axis produces a field of $20\,\si{\micro\tesla}$, comparable to the
Earth's field; the compass-deflection experiment that Oersted
performed in 1820 is recoverable on a bench with no special equipment.

A long solenoid with $n$ turns per unit length carrying current $I$
produces an axial field
\[
B = \mu_{0} n I
\]
inside, uniform across the cross-section in the limit of an infinite
solenoid; outside, the field is negligible compared to the interior.
A solenoid with $n = 1000\,\si{\per\meter}$ at $I = 10\,\si{\ampere}$
produces $B = 12.6\,\si{\milli\tesla}$, a useful field for laboratory
purposes.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-wire-bfield.tex}

The same Biot-Savart integral that yields these symmetric closed forms
also handles geometries without symmetry, where the only route is direct
quadrature. The script \repopath{code/biot\_savart\_segment.py} integrates
$\dd\vect{B} = (\mu_0 I/4\pi)\,\dd\vect{\ell}\times\unitvec{r}/r^{2}$
along a finite straight segment and along the four sides of a square
loop; it recovers the infinite-wire result $\mu_0 I/(2\pi r)$ as the
segment lengthens and the centre-of-loop value $2\sqrt{2}\,\mu_0 I/
(\pi a)$ for a square of side $a$. The lesson of the symmetric cases is
that they are the exceptions; most real coils are integrated
numerically.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-maxwell-summary.tex}

With the four equations stated, their differential forms collected in
\cref{fig:vol04ch08:maxwell-summary} read as two balances and two
transport laws. The remaining sections draw the consequences: the energy
the fields carry, the conditions they obey at material boundaries, and
the failure modes they produce in real systems. Before those
consequences, the two forms of each equation, the potentials that
generate the fields, and the regimes in which the full equations collapse
to circuit laws deserve a careful pass, because each move was used above
without its justification spelled out.

\section{The mathematical structure of the four equations}\pathtag{standard}

The four equations appeared twice each, once as an integral over a finite
region and once as a differential statement at a point, with the remark
that one follows from the other by a theorem of vector calculus. The
equivalence is exact, and the two theorems that carry it, the divergence
theorem and Stokes's theorem of Volume II, are worth applying in full,
because the integral form is the one an engineer measures and the
differential form is the one a solver integrates.

\subsection{Divergence theorem: the two balance laws}

The divergence theorem states that the flux of any vector field
$\vect{F}$ out through a closed surface $\partial V$ equals the integral
of its divergence over the enclosed volume,
\[
\oint_{\partial V}\vect{F}\cdot\dd\vect{A} = \int_{V}\divg\vect{F}\,\dd V.
\]
Apply it to Gauss's law. The integral form sets the outward flux of
$\vect{E}$ equal to the enclosed charge over $\varepsilon_0$,
\[
\oint_{\partial V}\vect{E}\cdot\dd\vect{A} = \frac{1}{\varepsilon_0}
\int_{V}\rho\,\dd V,
\]
and the divergence theorem rewrites the left side as a volume integral,
\[
\int_{V}\divg\vect{E}\,\dd V = \int_{V}\frac{\rho}{\varepsilon_0}\,\dd V.
\]
The two volume integrals are equal for every region $V$, including
regions made arbitrarily small around any point. Two continuous functions
whose integrals agree over every region agree pointwise, so the
integrands themselves are equal,
\[
\divg\vect{E} = \frac{\rho}{\varepsilon_0},
\]
the differential Gauss's law. The step that the integral form holds for
\emph{every} volume is what licenses dropping the integral signs; a
relation true only for one particular surface would yield no pointwise
law. The identical argument on $\oint\vect{B}\cdot\dd\vect{A} = 0$ gives
$\divg\vect{B} = 0$, since the enclosed magnetic charge is zero for every
volume. The two balance equations are the two applications of one
theorem.

\subsection{Stokes's theorem: the two transport laws}

Stokes's theorem states that the circulation of a vector field around a
closed curve $C$ equals the flux of its curl through any surface $S$ that
$C$ bounds,
\[
\oint_{C}\vect{F}\cdot\dd\vect{\ell} = \int_{S}(\curl\vect{F})\cdot
\dd\vect{A}.
\]
Apply it to Faraday's law. The integral form sets the circulation of
$\vect{E}$ around $C$ equal to minus the rate of change of flux through
$S$,
\[
\oint_{C}\vect{E}\cdot\dd\vect{\ell} = -\dv{}{t}\int_{S}\vect{B}\cdot
\dd\vect{A}.
\]
Stokes's theorem rewrites the left side as a surface integral of the
curl, and for a fixed surface the time derivative passes inside the
integral on the right,
\[
\int_{S}(\curl\vect{E})\cdot\dd\vect{A} = -\int_{S}\pdv{\vect{B}}{t}\cdot
\dd\vect{A}.
\]
The equality holds for every surface $S$ spanning every curve $C$,
including vanishingly small surfaces around any point with any
orientation of $\dd\vect{A}$. Two vector fields whose flux agrees through
every oriented surface element are equal, so the integrands match,
\[
\curl\vect{E} = -\pdv{\vect{B}}{t},
\]
the differential Faraday law. The same theorem on the
Amp\`ere-Maxwell circulation gives $\curl\vect{B} = \mu_0\vect{J} +
\mu_0\varepsilon_0\,\partial\vect{E}/\partial t$. The two transport
equations are the two applications of the second theorem, and the four
differential laws are now derived rather than asserted. The
correspondence is collected in
\cref{tab:vol04ch08:integral-differential}: each balance law is a
divergence theorem, each transport law a Stokes's theorem, and the
integral form is what a flux meter or a voltmeter loop reads while the
differential form is what a field solver advances.

\begin{table}[htbp]
\centering
\caption[Integral and differential forms]{The two forms of each Maxwell
equation and the theorem that connects them. The integral form is the
measured statement over a finite region; the differential form is the
pointwise statement a solver integrates. The balance laws use the
divergence theorem; the transport laws use Stokes's theorem.}
\label{tab:vol04ch08:integral-differential}
\small
\begin{tabular}{@{}lll@{}}
\toprule
Integral form & Differential form & Theorem \\
\midrule
$\oint\vect{E}\cdot\dd\vect{A} = Q_{\text{enc}}/\varepsilon_0$ &
$\divg\vect{E} = \rho/\varepsilon_0$ & divergence \\
$\oint\vect{B}\cdot\dd\vect{A} = 0$ &
$\divg\vect{B} = 0$ & divergence \\
$\oint\vect{E}\cdot\dd\vect{\ell} = -\dot\Phi_B$ &
$\curl\vect{E} = -\partial_t\vect{B}$ & Stokes \\
$\oint\vect{B}\cdot\dd\vect{\ell} = \mu_0(I + \varepsilon_0\dot\Phi_E)$ &
$\curl\vect{B} = \mu_0\vect{J} + \mu_0\varepsilon_0\partial_t\vect{E}$ &
Stokes \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Charge conservation is contained in Maxwell's equations}

The continuity equation $\partial\rho/\partial t + \divg\vect{J} = 0$
entered the chapter as a separate accounting statement. It is not
separate: it is a consequence of two of Maxwell's equations and need not
be imposed by hand. Take the divergence of the differential
Amp\`ere-Maxwell law,
\[
\divg(\curl\vect{B}) = \mu_0\divg\vect{J} + \mu_0\varepsilon_0\,
\divg\pdv{\vect{E}}{t}.
\]
The left side vanishes identically, because the divergence of any curl is
zero. On the right, the divergence and the time derivative commute, and
Gauss's law replaces $\divg\vect{E}$ with $\rho/\varepsilon_0$,
\[
0 = \mu_0\divg\vect{J} + \mu_0\varepsilon_0\pdv{}{t}\!\left(
\frac{\rho}{\varepsilon_0}\right) = \mu_0\!\left(\divg\vect{J} +
\pdv{\rho}{t}\right).
\]
Dividing by $\mu_0$ returns continuity exactly. Charge conservation is
built into the structure: any field configuration that satisfies Gauss's
law and the Amp\`ere-Maxwell law automatically conserves charge, which is
the deeper reason Maxwell's displacement term was forced rather than
chosen. The same calculation read backward, demanding continuity and
asking what term must be added to the bare Amp\`ere law to supply it, is
the derivation the exercises ask the reader to carry out.

\subsection{Potentials and gauge freedom}

The two homogeneous equations, $\divg\vect{B} = 0$ and
$\curl\vect{E} = -\partial_t\vect{B}$, are solved identically by writing
the fields in terms of potentials. Because the divergence of a curl
vanishes, $\divg\vect{B} = 0$ is satisfied for \emph{any} vector
potential $\vect{A}$ by setting
\[
\vect{B} = \curl\vect{A}.
\]
Substituting this into Faraday's law gives $\curl\vect{E} =
-\partial_t(\curl\vect{A}) = -\curl(\partial_t\vect{A})$, so
$\curl(\vect{E} + \partial_t\vect{A}) = \vect{0}$. A curl-free field is
the gradient of a scalar, so $\vect{E} + \partial_t\vect{A} =
-\grad\phi$ for some scalar potential $\phi$, the minus sign chosen so
that $\phi$ reduces to the electrostatic potential when nothing changes
in time,
\[
\vect{E} = -\grad\phi - \pdv{\vect{A}}{t}.
\]
The static field $\vect{E} = -\grad\phi$ of Chapter 9 is the special case
with $\partial_t\vect{A} = \vect{0}$. Two of the four equations are now
solved automatically by the very form of the potentials, and the
remaining two, Gauss and Amp\`ere-Maxwell, become equations for $\phi$
and $\vect{A}$.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-potentials-gauge.tex}

The potentials are not unique. Adding the gradient of any scalar function
$\chi(\vect{r}, t)$ to $\vect{A}$ leaves $\vect{B}$ unchanged, since
$\curl(\grad\chi) = \vect{0}$; to keep $\vect{E}$ unchanged as well, the
scalar potential must change by $-\partial_t\chi$ at the same time. The
pair of transformations
\[
\vect{A}\to\vect{A} + \grad\chi, \qquad \phi\to\phi - \pdv{\chi}{t}
\]
is the \engterm{gauge transformation}, and the invariance of the fields
under it is \engterm{gauge freedom}, the relation drawn in
\cref{fig:vol04ch08:potentials-gauge}. Because only $\vect{E}$ and
$\vect{B}$ are measured, the value of $\chi$ is free, and the engineer
fixes it by a \engterm{gauge condition} chosen to simplify the equations
that remain.

Two choices dominate. The \engterm{Coulomb gauge} sets
$\divg\vect{A} = 0$. With it, Gauss's law becomes Poisson's equation
$\laplacian\phi = -\rho/\varepsilon_0$, whose solution is the
instantaneous Coulomb potential of every charge, the form that makes the
scalar potential easy in quasi-static and circuit problems. The
\engterm{Lorenz gauge} sets $\divg\vect{A} + \mu_0\varepsilon_0\,
\partial\phi/\partial t = 0$. With it the two remaining Maxwell equations
become symmetric wave equations,
\[
\laplacian\phi - \mu_0\varepsilon_0\frac{\partial^{2}\phi}{\partial t^{2}}
= -\frac{\rho}{\varepsilon_0}, \qquad
\laplacian\vect{A} - \mu_0\varepsilon_0\frac{\partial^{2}\vect{A}}{
\partial t^{2}} = -\mu_0\vect{J},
\]
each a source driving a wave at speed $c = 1/\sqrt{\mu_0\varepsilon_0}$.
The radiation that a transmitter launches and the fields of a moving
charge both fall out of these two equations once the sources are
specified, and the Lorenz gauge is the natural one for radiation because
it puts the source and the wave on opposite sides of the same operator
\cite[ch.~10]{text:griffiths2017}. The choice of gauge changes the
potentials but never the fields, so any measurable prediction is
independent of it; the gauge is a coordinate freedom in the description,
not a fact about the world.

\begin{mastery}
The Lorenz-gauge wave equations for $\phi$ and $\vect{A}$ have a solution
that reads causality straight off the mathematics. The potential at a point
$\vect{r}$ and time $t$ is the sum over all sources, each evaluated not at
the present instant but at the earlier \engterm{retarded time} $t_r = t -
|\vect{r} - \vect{r}'|/c$, one light-travel time in the past,
\[
\phi(\vect{r},t) = \frac{1}{4\pi\varepsilon_0}\int
\frac{\rho(\vect{r}', t_r)}{|\vect{r} - \vect{r}'|}\,\dd V', \qquad
\vect{A}(\vect{r},t) = \frac{\mu_0}{4\pi}\int
\frac{\vect{J}(\vect{r}', t_r)}{|\vect{r} - \vect{r}'|}\,\dd V'.
\]
Each source element contributes what it was doing when the light now
arriving left it, so a change in a charge or a current is felt at a distant
point only after the delay light needs to cross the gap, never sooner. The
retarded solution is the reason a field carries information at exactly $c$
and no faster, the reason an antenna's radiation is set by what its charges
did nanoseconds ago rather than now, and the formal statement that the
displacement current Maxwell added did more than balance the books: it gave
the fields a finite propagation speed and made them carriers of energy and
signal in their own right, decoupled from their sources once launched. The
quasi-static approximation of the next subsection is precisely the limit in
which the retardation $|\vect{r} - \vect{r}'|/c$ is negligible against the
timescale of the source, so the retarded time collapses to the present and
the potentials adjust as if instantaneously. When that collapse is not
legitimate, the retarded potentials are the honest tool, and the whole of
radiation and antenna theory is their evaluation for a specified source.
\end{mastery}

\subsection{Quasi-static regimes: when the wave terms drop}

The full equations couple $\vect{E}$ and $\vect{B}$ through their time
derivatives, and that coupling is what produces waves. Across most of
engineering the coupling is negligible, and dropping one or both time
derivatives reduces the four equations to the laws of circuits and
magnetostatics. The condition for the reduction is geometric: a system of
size $L$ operating at frequency $f$ is quasi-static when $L$ is much
smaller than the wavelength $\lambda = c/f$, so that a field change at one
end reaches the other end in a time short compared with a period. The
ratio $L/\lambda = Lf/c$ is the parameter, and the regimes split as in
\cref{fig:vol04ch08:quasistatic-regimes}.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-quasistatic-regimes.tex}

When $L\ll\lambda$, propagation delay across the device is negligible and
the fields adjust as if instantaneously. Two sub-regimes follow from
which time derivative survives. In the \engterm{electro-quasi-static}
regime the displacement current matters but induction does not, so
$\curl\vect{E}\approx\vect{0}$ and the electric field is still the
gradient of a potential; this is the regime of capacitors, electrometers,
and high-impedance sensing, where charge redistributes but magnetic
induction is ignorable. In the \engterm{magneto-quasi-static} regime
induction matters but displacement current does not, so
$\curl\vect{B}\approx\mu_0\vect{J}$ holds while Faraday's law is kept in
full; this is the regime of transformers, inductors, eddy-current brakes,
and the skin effect, where currents and their induced EMFs dominate and
the displacement term is dropped. The lumped-element circuit theory of
Chapter 9, with its capacitances, inductances, and resistances, is the
quasi-static limit of Maxwell's equations on a system small compared with
a wavelength; Kirchhoff's voltage law is Faraday's law with negligible
enclosed flux, and Kirchhoff's current law is continuity with negligible
charge accumulation at a node.

The boundary is not academic. A circuit board trace that behaves as a
simple wire at audio frequency becomes a transmission line at a
gigahertz, because its length crosses from $L\ll\lambda$ to
$L\sim\lambda$ as the frequency rises by six orders of magnitude. The
practical rule places the boundary near $L = \lambda/10$, where the
propagation delay reaches a tenth of a period and reflections begin to
matter; below it the lumped model is safe, above it the full-wave
treatment of Chapter 10 is required. The same physical board is a circuit
in one regime and an antenna in the other, and the engineer's first
question about any electromagnetic system is which side of the line it
sits on.

\section{Electromagnetic energy and the Poynting vector}\pathtag{core}

The energy density stored in electric and magnetic fields is
\[
u_{e} = \tfrac{1}{2}\varepsilon_{0}|\vect{E}|^{2}, \quad u_{m} =
\tfrac{1}{2\mu_{0}}|\vect{B}|^{2},
\]
both in $\si{\joule\per\meter\cubed}$. A parallel-plate capacitor at
$1\,\si{\kilo\volt}$ with $1\,\si{\milli\meter}$ plate spacing has
$E = 10^{6}\,\si{\volt\per\meter}$ and $u_{e} = 4.4\,\si{\joule\per
\meter\cubed}$; the energy in a $1\,\si{\centi\meter\cubed}$ volume is
about $4\,\si{\micro\joule}$. A MRI magnet at $3\,\si{\tesla}$ has
$u_{m} = 3.6\,\si{\mega\joule\per\meter\cubed}$; the energy stored in a
$1\,\si{\meter\cubed}$ bore is $3.6\,\si{\mega\joule}$, several times
the kinetic energy of a small car at $100\,\si{\kilo\meter\per\hour}$.
The magnetic field is the heavier energy reservoir for the same volume
once $B$ exceeds the order of $\sqrt{\mu_{0}\varepsilon_{0}}\,E$,
which is to say $E/B \sim c$, the speed of light.

Both energy densities are derivable from the work done in establishing
the field, and the derivation is worth carrying out because it locates
the factor of one half that the lumped-element energies
$\tfrac12 CV^{2}$ and $\tfrac12 LI^{2}$ share with the field densities.
Consider first the parallel-plate capacitor of plate area $A$ and gap
$d$, charged to a final voltage $V$ holding a uniform interior field
$E = V/d$ and surface charge $\sigma = \varepsilon_{0} E$ on each plate.
Bring the charge on in increments. When the plate already carries
surface charge $\sigma'$, the field is $E' = \sigma'/\varepsilon_{0}$
and the work to move an extra $\dd\sigma'$ per unit area across the gap
$d$ is $E' d\,\dd\sigma' = (d/\varepsilon_{0})\sigma'\,\dd\sigma'$.
Integrating from zero to the final $\sigma = \varepsilon_{0} E$,
\[
\frac{W}{A} = \frac{d}{\varepsilon_{0}}\int_{0}^{\varepsilon_{0}E}
\sigma'\,\dd\sigma' = \frac{d}{\varepsilon_{0}}\cdot
\frac{(\varepsilon_{0}E)^{2}}{2} = \tfrac{1}{2}\varepsilon_{0}E^{2}\,d.
\]
The work per unit area divided by the gap $d$ is the work per unit
volume, and the field fills exactly the volume $Ad$ between the plates,
so the stored energy density is
\[
u_{e} = \frac{W}{Ad} = \tfrac{1}{2}\varepsilon_{0}E^{2},
\]
recovering the stated form from the work integral alone. The factor of
one half is the average field seen by the charge as it is carried over,
not the final field; it is the area of the triangle under the
constitutive line $D = \varepsilon_{0}E$ in the left panel of
\cref{fig:vol04ch08:field-energy-density}. In a linear dielectric the
same argument with $D = \varepsilon E$ gives $u_{e} = \tfrac12 \vect{E}
\cdot\vect{D}$, the form valid in matter.

The magnetic density follows from the dual argument on a long solenoid.
A solenoid of $n$ turns per unit length carrying current $I$ holds an
interior field $B = \mu_{0} n I$ and stores $U_{m} = \tfrac12 L I^{2}$
with inductance per unit length $L' = \mu_{0} n^{2}\pi a^{2}$ from the
worked example of Section 8.8. The stored energy per unit length is
$\tfrac12 L' I^{2} = \tfrac12\mu_{0} n^{2}\pi a^{2} I^{2}$, and the
field fills the interior cross-section $\pi a^{2}$, so the energy per
unit volume is
\[
u_{m} = \frac{\tfrac12\mu_{0} n^{2}\pi a^{2} I^{2}}{\pi a^{2}} =
\tfrac{1}{2}\mu_{0} n^{2} I^{2} = \frac{(\mu_{0} n I)^{2}}{2\mu_{0}} =
\frac{B^{2}}{2\mu_{0}},
\]
the stated form, again as the triangle under the line $B = \mu_{0} H$ in
the right panel of \cref{fig:vol04ch08:field-energy-density}. The two
densities are not separate facts; they are the same accounting applied
once to charge moved against a field and once to current driven against
a back-EMF, and the factor of one half in each is the linear build-up of
the field from zero. The Poynting theorem below makes this energy
mobile: it tells us how the stored density flows from one place to
another.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-field-energy-density.tex}

The rate at which electromagnetic energy flows through a surface is the
flux of the \engterm{Poynting vector},
\[
\vect{S} = \frac{1}{\mu_{0}}\vect{E}\times \vect{B},
\]
in $\si{\watt\per\meter\squared}$. The Poynting theorem, derived by
taking $\vect{E}\cdot$ (the Amp\`ere-Maxwell equation) minus
$\vect{B}\cdot$ (Faraday's equation) and integrating, states
\[
\pdv{(u_{e}+u_{m})}{t} + \divg \vect{S} = -\vect{J}\cdot \vect{E},
\]
the local energy balance: the rate of change of stored field energy
plus the divergence of the Poynting vector equals the negative of the
work done by the field on the current. The right-hand side is the
mechanism by which field energy becomes thermal energy in a resistor:
$\vect{J}\cdot \vect{E}$ is the volumetric power density dissipated.

\begin{principle}[Energy travels in the fields, not in the wire]
For a coaxial cable carrying DC current from a source to a load, the
electric field points radially inward, the magnetic field circulates
azimuthally, and the Poynting vector $\vect{E}\times \vect{B}/\mu_{0}$
points along the cable's axis from source to load. Integrating over
the cable's cross-section gives a power flux equal to $VI$, the
familiar product of voltage and current. The energy is carried in the
space between the conductors, not in the conductors themselves. This
is more than a curiosity: it is the working content of transmission-
line theory in Chapter 10, where impedance and standing waves emerge
from the field's path between conductors.
\end{principle}

A plane electromagnetic wave in vacuum has $|\vect{E}|/|\vect{B}| = c
= 1/\sqrt{\mu_{0}\varepsilon_{0}} = 2.998\times 10^{8}\,\si{\meter\per
\second}$, and equal contributions to $u_{e}$ and $u_{m}$. The
time-averaged Poynting vector magnitude for a sinusoidal plane wave
of peak electric-field amplitude $E_{0}$ is
\[
\langle S\rangle = \frac{1}{2}\frac{E_{0}^{2}}{\mu_{0} c} =
\frac{1}{2}\varepsilon_{0} c E_{0}^{2}.
\]
Solar irradiance at the top of Earth's atmosphere is about $1361\,\si{
\watt\per\meter\squared}$ (the \engterm{solar constant}, current as
of 2025); equating this to $\langle S\rangle$ gives a peak field of
$E_{0} \approx 1000\,\si{\volt\per\meter}$. The same calculation drives
the design of radio receivers (Chapter 10) and photovoltaic cells
(Chapter 12). The script \repopath{code/plane\_wave.py} carries the
conversion both ways, from irradiance to peak field amplitudes and back,
together with the time-averaged energy density and the radiation
pressure on an absorbing surface; at the solar constant it returns
$E_{0} = 1013\,\si{\volt\per\meter}$, $B_{0} = 3.4\,\si{\micro\tesla}$,
and a radiation pressure of $4.5\,\si{\micro\pascal}$.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-em-wave.tex}

\begin{mastery}
The wave itself falls out of the two curl equations in vacuum, where
$\rho = 0$ and $\vect{J} = \vect{0}$. Take the curl of Faraday's law,
$\curl(\curl\vect{E}) = -\partial(\curl\vect{B})/\partial t$, and
substitute the Amp\`ere-Maxwell law $\curl\vect{B} = \mu_0\varepsilon_0\,
\partial\vect{E}/\partial t$ on the right. The vector identity
$\curl(\curl\vect{E}) = \grad(\divg\vect{E}) - \laplacian\vect{E}$ with
$\divg\vect{E} = 0$ collapses the left side, leaving
\[
\laplacian\vect{E} = \mu_0\varepsilon_0\frac{\partial^{2}\vect{E}}{
\partial t^{2}},
\]
the wave equation with speed $c = 1/\sqrt{\mu_0\varepsilon_0}$. The
identical manipulation on $\vect{B}$ gives the same equation. The speed
came out of two laboratory constants measured in electrostatics and
magnetostatics, and it matched the measured speed of light; that
coincidence is the content of the chapter's epigraph. The same
second-order wave operator governed the vibrating string in Volume II
and recurs in transmission-line theory in Chapter 10.
\end{mastery}

\section{Boundary conditions and material laws}\pathtag{standard}

Maxwell's equations in their bare form describe fields in vacuum. In
material media, the bound charges and bound currents of the material
modify the fields. The book's working notation introduces two
auxiliary fields: the \engterm{electric displacement} $\vect{D}$ and
the \engterm{magnetic field intensity} $\vect{H}$, defined by
\[
\vect{D} = \varepsilon_{0}\vect{E} + \vect{P}, \quad \vect{H} =
\frac{1}{\mu_{0}}\vect{B} - \vect{M},
\]
where $\vect{P}$ is the polarisation density (dipole moment per unit
volume) and $\vect{M}$ is the magnetisation (magnetic moment per unit
volume). The Maxwell equations in matter become
\[
\divg \vect{D} = \rho_{f}, \quad \divg \vect{B} = 0,
\]
\[
\curl \vect{E} = -\pdv{\vect{B}}{t}, \quad \curl \vect{H} = \vect{J}_{f}
+ \pdv{\vect{D}}{t},
\]
where the subscript $f$ denotes free (as opposed to bound) charges and
currents. For linear isotropic materials,
\[
\vect{D} = \varepsilon \vect{E}, \quad \vect{B} = \mu \vect{H},
\quad \vect{J}_{f} = \sigma \vect{E},
\]
introducing the \engterm{permittivity} $\varepsilon$, the
\engterm{permeability} $\mu$, and the \engterm{conductivity} $\sigma$.
The relative quantities are $\varepsilon_{r} = \varepsilon/\varepsilon
_{0}$ and $\mu_{r} = \mu/\mu_{0}$. Water has $\varepsilon_{r} \approx
80$ at low frequencies; ferrite has $\mu_{r}$ from a few to $10\,000$;
copper has $\sigma \approx 5.96\times 10^{7}\,\si{\siemens\per\meter}$;
glass at room temperature has $\sigma \approx 10^{-12}\,\si{\siemens
\per\meter}$.

The three parameters span the working range of every material an
electrical engineer selects. Conductivity ranges over twenty-two orders
of magnitude from copper to glass, the widest spread of any material
property in engineering, which is why the same physics describes both a
wire and its insulation and why the boundary between them is so sharp. A
material is called a conductor or an insulator by which term of the
Amp\`ere-Maxwell law dominates at the frequency of interest: the
conduction current $\sigma\vect{E}$ or the displacement current
$\varepsilon\,\partial\vect{E}/\partial t$. Their ratio is $\sigma/
(\omega\varepsilon)$, the \engterm{loss tangent}; a material with loss
tangent far above one behaves as a conductor, far below one as a
dielectric, and the same material can cross from one regime to the other
as frequency rises, which is why seawater conducts at audio frequency
and is nearly transparent to it at optical frequency. The selection of a
dielectric for a capacitor, an insulator for a cable, or a shield for an
enclosure is the selection of a point in the
$(\varepsilon, \mu, \sigma)$ space against the frequency the device must
work at.

The third Maxwell-in-matter equation is Ohm's law in field form:
$\vect{J} = \sigma \vect{E}$ is the constitutive equation of a linear
conductor, and the lumped-element relation $V = IR$ derived in
Chapter 9 is its integral over a wire's cross-section and length.

At a sharp boundary between materials $1$ and $2$ with unit normal
$\unitvec{n}$ from $1$ to $2$, the field components obey:
\begin{itemize}[leftmargin=*]
  \item $D_{2n} - D_{1n} = \sigma_{f}$ (normal $\vect{D}$ jumps by the
    free surface charge density).
  \item $B_{2n} - B_{1n} = 0$ (normal $\vect{B}$ is continuous).
  \item $\vect{E}_{2t} - \vect{E}_{1t} = \vect{0}$ (tangential
    $\vect{E}$ is continuous).
  \item $\unitvec{n}\times(\vect{H}_{2} - \vect{H}_{1}) = \vect{K}_{f}$
    (tangential $\vect{H}$ jumps by the free surface current
    density).
\end{itemize}
For a perfect conductor, $\vect{E}$ and $\vect{B}$ inside are zero in
steady state, so at the surface $\vect{E}$ is normal to the surface
and $\vect{B}$ is tangential. The four boundary conditions are the
interface archetype: every reflection, refraction, waveguide mode, and
transmission-line discontinuity in Chapter 10 is solved by enforcing
them.

\begin{mastery}
A conductor is not a perfect conductor at high frequency. The
constitutive law $\vect{J} = \sigma\vect{E}$ combined with the
Amp\`ere-Maxwell and Faraday laws gives, inside a good conductor where
displacement current is negligible, the diffusion equation
$\laplacian\vect{E} = \mu\sigma\,\partial\vect{E}/\partial t$. A
sinusoidal field at angular frequency $\omega$ decays into the conductor
as $\exp(-x/\delta)$ with the \engterm{skin depth}
\[
\delta = \sqrt{\frac{2}{\mu\sigma\omega}}.
\]
For copper at $50\,\si{\hertz}$ the skin depth is about $9\,\si{\milli
\meter}$; at $1\,\si{\mega\hertz}$ it is $65\,\si{\micro\meter}$; at
$1\,\si{\giga\hertz}$ it is about $2\,\si{\micro\meter}$. Current at
radio frequency rides on the surface, so the usable cross-section of a
wire shrinks with frequency, its effective resistance rises, and the
designer plates conductors with thin layers of high-conductivity metal
rather than paying for solid bulk. The same skin depth sets the
thickness of electromagnetic shielding: a shield works once it is
several skin depths thick at the frequency it must block.
\end{mastery}

\section{Worked examples: solenoid, parallel-plate capacitor, coaxial cable}\pathtag{standard}

A long solenoid with $n$ turns per unit length, radius $a$, carrying
current $I$, has an interior field $\vect{B} = \mu_{0} n I\unitvec{z}$
and a negligible exterior field. Its self-inductance per unit length
is computed from the flux linkage per ampere:
\[
\Phi_{\text{per turn}} = B\pi a^{2} = \mu_{0} n I \pi a^{2},
\]
and the linkage per unit length is $n$ turns per metre times
$\Phi_{\text{per turn}}$:
\[
L' = \frac{N\Phi_{\text{per turn}}}{I\ell} = \mu_{0} n^{2} \pi a^{2}.
\]
A solenoid with $n = 1000\,\si{\per\meter}$ and $a = 1\,\si{\centi
\meter}$ has $L' = 4\pi\times 10^{-7}\times 10^{6}\times \pi\times
10^{-4} \approx 0.4\,\si{\milli\henry\per\meter}$. The energy stored
per unit length is $\tfrac{1}{2}L' I^{2}$; at $I = 10\,\si{\ampere}$,
$U/\ell \approx 20\,\si{\milli\joule\per\meter}$.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-parallel-plate.tex}

A parallel-plate capacitor with plate area $A$, separation $d$, and
dielectric permittivity $\varepsilon$ has a uniform interior field
$E = V/d$ when held at voltage $V$. The charge on a plate, by Gauss's
law on a pillbox spanning the plate, is $Q = \varepsilon E A =
\varepsilon V A/d$, giving the capacitance
\[
C = \frac{Q}{V} = \frac{\varepsilon A}{d}.
\]
A $1\,\si{\centi\meter\squared}$ plate with $1\,\si{\milli\meter}$
air gap has $C = 8.85\times 10^{-13}\,\si{\farad} = 0.885\,\si{\pico
\farad}$; the picofarad is the natural scale of bench-built
capacitors, and only with thin dielectrics ($d \ll 1\,\si{\milli
\meter}$) and high $\varepsilon_{r}$ (ceramics, electrolytes) do
practical capacitances of microfarads and millifarads become
achievable. The dielectric constants that make these high capacitances
practical are collected in \repopath{data/dielectric\_materials.csv},
which lists relative permittivity and dielectric strength from vacuum
through air, polyethylene, and the class-2 ceramics whose
$\varepsilon_r$ reaches several thousand. The idealised formula
$C = \varepsilon A/d$ omits the fringing field at the plate edges; the
finite-difference solver \repopath{code/laplace\_capacitor.py} relaxes
Laplace's equation on a grid for finite plates and reports the fringing
correction as a function of plate aspect ratio, the quantity the
simulation exercise asks the reader to characterise.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-coax-cross-section.tex}

A coaxial cable of inner radius $a$, outer radius $b$, with the
dielectric of permittivity $\varepsilon$ between, carries a linear
charge $\lambda$ on the inner conductor (and $-\lambda$ on the outer).
Gauss's law on a cylindrical Gaussian surface gives
\[
E(r) = \frac{\lambda}{2\pi \varepsilon r}, \quad a<r<b.
\]
The potential difference between the conductors is
\[
V = -\int_{b}^{a} E\,\dd r = \frac{\lambda}{2\pi \varepsilon}\ln(b/a),
\]
and the capacitance per unit length is
\[
C' = \frac{\lambda}{V} = \frac{2\pi \varepsilon}{\ln(b/a)}.
\]
A polyethylene-dielectric cable with $\varepsilon_{r} = 2.25$, $b/a =
3$, has $C' = 2\pi\varepsilon_{0}\times 2.25/\ln 3 \approx 114\,\si{
\pico\farad\per\meter}$, the standard $50\,\si{\ohm}$ RG-58 cable.
The inductance per unit length, computed from Amp\`ere's law and the
flux linkage between the conductors, is
\[
L' = \frac{\mu}{2\pi}\ln(b/a),
\]
giving a characteristic impedance
\[
Z_{0} = \sqrt{L'/C'} = \frac{1}{2\pi}\sqrt{\mu/\varepsilon}\,\ln(b/a).
\]
For air dielectric, $Z_{0} \approx 60\,\ln(b/a)$; for polyethylene,
$Z_{0} \approx 40\,\ln(b/a)$, recovering $50\,\si{\ohm}$ at $b/a
\approx 3$. The characteristic impedance is the load that, when
attached at the cable's far end, absorbs all incident power; mismatched
loads produce reflections, treated in Chapter 10. The script
\repopath{code/coax\_parameters.py} evaluates $C'$, $L'$, $Z_0$, and the
propagation velocity for any geometry and dielectric; for the RG-58
parameters it returns $Z_0 \approx 47\,\si{\ohm}$ and a propagation
velocity of $0.67c$, the velocity factor printed on the cable's
datasheet.

\subsection{Energy stored, force, and the factor of one half}

Three worked quantities recur across electromagnetic design, and each
carries a factor of one half that is the common source of error. We
collect them on a single capacitor and inductor so the pattern is
visible.

The energy stored in a capacitor charged to voltage $V$ is built up by
moving charge against the rising field. Bring an increment $\dd q$ onto
a plate already at potential $v = q/C$; the work is $v\,\dd q$, and the
total work to reach charge $Q$ is
\[
U_e = \int_{0}^{Q} \frac{q}{C}\,\dd q = \frac{Q^{2}}{2C} =
\tfrac{1}{2}CV^{2} = \tfrac{1}{2}QV.
\]
The factor of one half is the average potential seen by the charge as
it is carried over, not the final potential. A $100\,\si{\micro\farad}$
capacitor at $50\,\si{\volt}$ stores $U_e = \tfrac12(10^{-4})(50)^{2} =
0.125\,\si{\joule}$, enough to deliver a sharp shock and to weld a
screwdriver tip that bridges its terminals.

The energy stored in an inductor carrying current $I$ is built up the
same way against the back-EMF. The voltage across an inductor is
$v = L\,\dd i/\dd t$, so the power delivered is $vi = Li\,\dd i/\dd t$,
and the energy to reach current $I$ is
\[
U_m = \int_{0}^{I} Li\,\dd i = \tfrac{1}{2}LI^{2}.
\]
The $1.6\,\si{\milli\henry}$ solenoid of the worked example, at
$3\,\si{\ampere}$, stores $U_m = \tfrac12(1.6\times 10^{-3})(3)^{2} =
7.2\,\si{\milli\joule}$. Interrupting that current quickly forces a
large $L\,\dd i/\dd t$ across the switch, the arc that destroys relay
contacts and the reason a freewheeling diode is placed across every
inductive load.

The force of attraction between the plates of a charged capacitor
follows from the energy. At fixed charge $Q$ the field energy is
$U_e = Q^{2}d/(2\varepsilon_0 A)$, since $C = \varepsilon_0 A/d$, and
the attractive force is the negative gradient of energy with separation,
\[
F = -\dv{U_e}{d} = \frac{Q^{2}}{2\varepsilon_0 A} = \tfrac{1}{2}
\varepsilon_0 E^{2}\,A,
\]
where $E = Q/(\varepsilon_0 A)$ is the interior field. The factor of one
half appears again, and for the same reason in disguise: the force on
one plate is set by the field of the \emph{other} plate alone, which is
$E/2$, so $F = QE/2$. The quantity $\tfrac12\varepsilon_0 E^{2}$ is the
electrostatic pressure, the same number as the electric energy density,
and it caps the field a capacitor can hold before the plates are pulled
into contact or the dielectric breaks down.

\subsection{Motional EMF: the rod, the rail, and the generator}

A conducting rod of length $\ell$ slides at speed $v$ along two parallel
rails in a uniform field $B$ perpendicular to the plane of the rails.
Each free charge in the rod feels the magnetic force $q\vect{v}\times
\vect{B}$, which drives positive charge to one end until the
electrostatic field it sets up balances the magnetic force. The
steady-state potential difference is the \engterm{motional EMF}
\[
\mathcal{E} = B\ell v,
\]
which is Faraday's law in its moving-circuit form: the rod sweeps area
$\ell v$ per unit time, so the flux through the rail circuit changes at
$B\ell v$. A rod $\ell = 0.5\,\si{\meter}$ long moving at $v = 10\,\si{
\meter\per\second}$ through $B = 0.2\,\si{\tesla}$ generates
$\mathcal{E} = (0.2)(0.5)(10) = 1.0\,\si{\volt}$. If the rails close
through a resistance $R = 2\,\si{\ohm}$, the current is $I = \mathcal{E}
/R = 0.5\,\si{\ampere}$, the dissipated power is $\mathcal{E}I =
0.5\,\si{\watt}$, and the retarding force on the rod is $F = BI\ell =
(0.2)(0.5)(0.5) = 0.05\,\si{\newton}$. The mechanical power needed to
hold the speed is $Fv = (0.05)(10) = 0.5\,\si{\watt}$, equal to the
electrical power dissipated. The bookkeeping closes: every watt of
electrical output is a watt of mechanical input, with the back-EMF
force as the coupling. A rotating coil in a field is the same device
wrapped into a circle, and the result $\mathcal{E} = N B A \omega
\sin\omega t$ is the alternating output of every rotating generator.

\subsection{Displacement current in a charging capacitor}

The displacement current is abstract until it is made to balance a real
conduction current. A parallel-plate capacitor of plate area $A$ and
gap $d$ is charged by a wire carrying conduction current $I$. The charge
on a plate is $Q = \varepsilon_0 E A$, so the field rises at $\dd E/\dd t
= I/(\varepsilon_0 A)$, and the displacement current between the plates
is
\[
I_d = \varepsilon_0\dv{}{t}\int_{S} E\,\dd A = \varepsilon_0 A
\dv{E}{t} = \varepsilon_0 A\cdot\frac{I}{\varepsilon_0 A} = I,
\]
exactly the conduction current in the wire. The continuity of the
Amp\`ere-Maxwell circulation is what demands this. Take the loop $C$ as
a circle of radius $r$ around the wire, and choose two surfaces it
bounds: a flat disc the wire pierces, and a bulging surface that passes
between the capacitor plates and is pierced by no wire. On the flat disc
the enclosed conduction current is $I$; on the bulging surface it is
zero, and only the displacement current $I_d = I$ supplies the same
circulation $\oint\vect{B}\cdot\dd\vect{\ell} = \mu_0 I$. For a charging
current of $I = 1\,\si{\ampere}$ into plates of radius $5\,\si{\centi
\meter}$, the magnetic field at the plate edge is $B = \mu_0 I_d/(2\pi r)
= (4\pi\times 10^{-7})(1)/(2\pi\times 0.05) = 4.0\,\si{\micro\tesla}$,
the same field a $1\,\si{\ampere}$ wire would produce at that radius. A
changing electric field is a source of magnetic field as surely as a
current is, which is the prediction that gave the chapter its wave.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-magnetic-circuit.tex}

\begin{mastery}
The reluctance analogy of \cref{fig:vol04ch08:magnetic-circuit} turns a
magnetic-circuit problem into a network problem the engineer already
knows how to solve. Around a closed core,
Amp\`ere's law gives $\oint\vect{H}\cdot\dd\vect{\ell} = NI$, the
\engterm{magnetomotive force} $\mathcal{F} = NI$. For a core of
cross-section $A$, mean path length $\ell$, and permeability $\mu$, the
flux $\Phi = BA = \mu H A$ is uniform around the loop, so
$\mathcal{F} = H\ell = (\Phi/\mu A)\,\ell$, which rearranges to
\[
\mathcal{F} = \Phi\,\mathcal{R}, \qquad \mathcal{R} =
\frac{\ell}{\mu A},
\]
the magnetic Ohm's law with reluctance $\mathcal{R}$ in the role of
resistance, flux $\Phi$ in the role of current, and MMF $\mathcal{F}$ in
the role of voltage. Reluctances in series add; an air gap of length
$g$ in a core of cross-section $A$ adds $\mathcal{R}_g = g/(\mu_0 A)$,
which dominates because $\mu_0 \ll \mu_{\text{core}}$: a gap a thousand
times shorter than the core path can carry most of the MMF. The analogy
has one limit worth stating. Reluctance is linear only while $\mu$ is
constant, and $\mu$ collapses as the core approaches saturation, so the
network model holds below the knee of the B-H curve and fails above it.
The case study that follows lives entirely below the knee, by design.
\end{mastery}

\section{An end-to-end case study: a small power transformer}\pathtag{standard}

The chapter's separate results converge when an engineer must design a
working magnetic device and account for every watt that enters it. We
take a single-phase $50\,\text{Hz}$ power transformer rated at
$S = 1\,\text{kVA}$, stepping $230\,\si{\volt}$ down to $24\,\si{\volt}$,
and carry it from the core flux through the turns count, the magnetising
current, the two loss mechanisms, the temperature rise, and the
efficiency, ending in numbers an engineer would sign. Every move of the
chapter appears once: Faraday's law sets the turns, Amp\`ere's law and
the reluctance analogy set the magnetising current, the B-H loop sets the
core loss, Ohm's law in the windings sets the copper loss, and the
Poynting balance accounts for the energy that does not reach the load.

\subsection{Core, flux, and the volts-per-turn rule}

The core is grain-oriented silicon steel with a cross-sectional area
$A_c = 16\,\si{\centi\meter\squared} = 1.6\times 10^{-3}\,\si{\meter
\squared}$ and a mean magnetic path length $\ell_c = 0.30\,\si{\meter}$.
The design places the peak flux density at $B_{\text{pk}} = 1.5\,\si{
\tesla}$, the operating point of \cref{fig:vol04ch08:bh-curve}, safely
below the saturation knee near $2.0\,\si{\tesla}$ so the permeability
stays high. The peak flux through the core is
\[
\Phi_{\text{pk}} = B_{\text{pk}} A_c = (1.5)(1.6\times 10^{-3}) =
2.4\times 10^{-3}\,\si{\weber}.
\]
For a sinusoidal supply at frequency $f$, Faraday's law applied to a
winding of $N$ turns gives a root-mean-square voltage $V_{\text{rms}} =
2\pi f N \Phi_{\text{pk}}/\sqrt{2} = 4.44\, f N \Phi_{\text{pk}}$, the
\engterm{transformer EMF equation}. The volts per turn is therefore
\[
\frac{V}{N} = 4.44\, f\,\Phi_{\text{pk}} = (4.44)(50)(2.4\times 10^{-3})
= 0.533\,\si{\volt}.
\]
The primary needs $N_1 = 230/0.533 = 432$ turns and the secondary
$N_2 = 24/0.533 = 45$ turns, rounding to convenient counts. The turns
ratio $N_1/N_2 = 432/45 = 9.6$ matches the voltage ratio $230/24 = 9.6$,
the consistency check the design must pass.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-bh-curve.tex}

\subsection{Magnetising current by the reluctance analogy}

Even with no load on the secondary, the primary draws a magnetising
current to establish the core flux. The reluctance of the core is
\[
\mathcal{R}_c = \frac{\ell_c}{\mu_r\mu_0 A_c} = \frac{0.30}{(5000)
(4\pi\times 10^{-7})(1.6\times 10^{-3})} = 2.98\times 10^{4}\,\si{\per
\henry},
\]
taking a relative permeability $\mu_r = 5000$ for the silicon steel at
the operating point. The peak magnetomotive force needed is $\mathcal{F}
= \Phi_{\text{pk}}\mathcal{R}_c = (2.4\times 10^{-3})(2.98\times 10^{4})
= 71.5\,\si{\ampere}$ (ampere-turns), so the peak magnetising current is
\[
I_{\mu,\text{pk}} = \frac{\mathcal{F}}{N_1} = \frac{71.5}{432} =
0.166\,\si{\ampere},
\]
about $0.12\,\si{\ampere}$ rms. Against a full-load primary current of
$S/V_1 = 1000/230 = 4.35\,\si{\ampere}$, the magnetising current is
under three percent, which confirms the core is operated where the
permeability is high. A gap or a lower-permeability core would raise the
reluctance and the magnetising current in direct proportion; this is the
reluctance analogy of the previous box used in anger.

\subsection{Core loss and copper loss}

Two loss mechanisms turn input power into heat. The core loss is the
hysteresis loop area of \cref{fig:vol04ch08:bh-curve} traversed $f$
times per second, plus eddy-current loss from the changing flux inducing
circulating currents in the steel. Grain-oriented silicon steel at
$B_{\text{pk}} = 1.5\,\si{\tesla}$ and $50\,\text{Hz}$ has a specific
core loss of about $1.1\,\si{\watt\per\kilogram}$. The core volume is
$A_c\ell_c = (1.6\times 10^{-3})(0.30) = 4.8\times 10^{-4}\,\si{\meter
\cubed}$, and at a steel density of $7650\,\si{\kilogram\per\meter
\cubed}$ the core mass is $3.7\,\si{\kilogram}$, giving
\[
P_{\text{core}} = (1.1)(3.7) \approx 4.0\,\si{\watt}.
\]
The lamination of the core into thin insulated sheets is what keeps the
eddy-current share of this number small; a solid core would dissipate
many times more. The copper loss is the $I^{2}R$ dissipation in the two
windings. Sizing each winding for a current density of $3\,\si{\ampere}
$ per square millimetre, the primary at $4.35\,\si{\ampere}$ uses
$1.45\,\si{\milli\meter\squared}$ wire and the secondary at
$S/V_2 = 41.7\,\si{\ampere}$ uses $13.9\,\si{\milli\meter\squared}$
wire. Estimating mean turn lengths of $0.18\,\si{\meter}$ and the
copper resistivity $1.72\times 10^{-8}\,\si{\ohm\meter}$ gives winding
resistances near $R_1 = 0.92\,\si{\ohm}$ and $R_2 = 0.0097\,\si{\ohm}$,
so
\[
P_{\text{cu}} = I_1^{2}R_1 + I_2^{2}R_2 = (4.35)^{2}(0.92) +
(41.7)^{2}(0.0097) \approx 17 + 17 = 34\,\si{\watt}
\]
at full load. The design splits the copper loss evenly between the two
windings, the standard choice that minimises total copper for a given
loss budget.

\subsection{Temperature rise, efficiency, and the verdict}

The total full-load loss is $P_{\text{loss}} = P_{\text{core}} +
P_{\text{cu}} = 4.0 + 34 = 38\,\si{\watt}$, dissipated from the
transformer's surface by natural convection and radiation. Taking an
external surface area of about $0.06\,\si{\meter\squared}$ and a
combined heat-transfer coefficient of $12\,\si{\watt\per\meter\squared}$
per kelvin (the convection result of Chapter 5), the steady temperature
rise is
\[
\Delta T = \frac{P_{\text{loss}}}{h A_s} = \frac{38}{(12)(0.06)} =
53\,\si{\kelvin}.
\]
A $53\,\si{\kelvin}$ rise over a $40\,\degC$ ambient reaches a
hot-spot temperature near $93\,\degC$, inside the rating of class-A
($105\,\degC$) winding insulation with margin. The efficiency at full
load is
\[
\eta = \frac{P_{\text{out}}}{P_{\text{out}} + P_{\text{loss}}} =
\frac{1000}{1000 + 38} = 0.963,
\]
about ninety-six percent, typical for a transformer of this rating; the
larger the unit, the higher the efficiency, because surface area and
copper volume scale more slowly than throughput. The loss split also
fixes the load at which efficiency peaks. Core loss is fixed and copper
loss rises as the square of load fraction $x$, so total loss is
$P_{\text{core}} + x^{2}P_{\text{cu}}$ and efficiency is greatest where
the two are equal, at $x = \sqrt{P_{\text{core}}/P_{\text{cu}}} =
\sqrt{4.0/34} = 0.34$, about a third of rated load. This is the content
of \cref{fig:vol04ch08:transformer-losses}: a transformer that spends
its life lightly loaded should be designed with its loss crossover at
that lighter load, which is why distribution-transformer standards
specify loss at a fraction of rating rather than at full rating alone.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-transformer-losses.tex}

The verdict is a device that meets its voltage ratio to better than a
percent, draws under three percent magnetising current, dissipates
$38\,\si{\watt}$ at full load for a $53\,\si{\kelvin}$ rise within its
insulation class, and runs at ninety-six percent efficiency. Every
number traces to one of the four equations: Faraday's law set the
turns, Amp\`ere's law and the reluctance analogy set the magnetising
current, the B-H loop set the core loss, and the resistive
$\vect{J}\cdot\vect{E}$ of the Poynting balance set the copper loss and
the heat that the temperature rise removes. The same procedure sizes an
inductor, a motor, or a magnetic actuator; only the boundary conditions
on the flux path change.

\section{A second case study: a coaxial transmission line end to end}\pathtag{standard}

The transformer carried field energy across a magnetic core; a
transmission line carries it along a guided path between two conductors,
and the design questions are different in kind. An engineer who must run
a $1\,\si{\giga\hertz}$ signal $30\,\si{\meter}$ from a transmitter to
an antenna, or feed pulsed power to a load, asks what impedance the
cable presents, how fast the signal travels, how much of it survives the
run, and how much voltage and power the cable can take before it fails.
We take a $50\,\si{\ohm}$ polyethylene-dielectric coaxial cable and
carry it from geometry through impedance, velocity, loss, and ratings to
numbers an engineer would put on a drawing. The per-unit-length results
of Section 8.8 are the starting point; this study turns them into a
specification \cite[ch.~2]{text:pozar-microwave}.

\subsection{Geometry, impedance, and velocity}

The cable has an inner conductor of radius $a = 0.45\,\si{\milli\meter}$
and an outer conductor of inner radius $b = 1.47\,\si{\milli\meter}$,
with a solid polyethylene dielectric of relative permittivity
$\varepsilon_r = 2.25$ filling the gap. The radius ratio is $b/a =
3.27$, so $\ln(b/a) = 1.184$. The per-unit-length parameters of
Section 8.8 give the capacitance and inductance directly:
\[
C' = \frac{2\pi\varepsilon_0\varepsilon_r}{\ln(b/a)} =
\frac{2\pi(8.85\times 10^{-12})(2.25)}{1.184} = 1.06\times 10^{-10}\,
\si{\farad\per\meter} = 106\,\si{\pico\farad\per\meter},
\]
\[
L' = \frac{\mu_0}{2\pi}\ln(b/a) = (2\times 10^{-7})(1.184) =
2.37\times 10^{-7}\,\si{\henry\per\meter} = 237\,\si{\nano\henry\per
\meter}.
\]
The characteristic impedance is the square root of their ratio,
\[
Z_0 = \sqrt{\frac{L'}{C'}} = \sqrt{\frac{2.37\times 10^{-7}}{1.06\times
10^{-10}}} = 47\,\si{\ohm},
\]
within the manufacturing tolerance of the nominal $50\,\si{\ohm}$ part.
A line terminated in this impedance absorbs all the power that reaches
it and reflects none; this is the content of the principle that opened
Section 8.6, made quantitative. The signal travels at the phase velocity
\[
v_p = \frac{1}{\sqrt{L'C'}} = \frac{c}{\sqrt{\varepsilon_r}} =
\frac{3.0\times 10^{8}}{\sqrt{2.25}} = 2.0\times 10^{8}\,\si{\meter\per
\second},
\]
two-thirds the speed of light, the \engterm{velocity factor} of $0.67$
printed on the cable's datasheet. The dielectric slows the wave because
the field must polarise the polyethylene as it passes, and the stored
energy in that polarisation is energy not available to advance the
wavefront. A $30\,\si{\meter}$ run therefore delays the signal by
$30/(2.0\times 10^8) = 150\,\si{\nano\second}$, a delay that matters
when the cable is part of a timing chain or a phased array.

\subsection{Attenuation: the skin-loss budget}

A real cable loses power along its length. Two mechanisms dissipate it:
the resistance of the conductors and the loss in the dielectric. The
conductor loss is set by the skin effect of the Section 8.7 mastery box.
At $1\,\si{\giga\hertz}$ the skin depth in copper is
\[
\delta = \sqrt{\frac{2}{\mu_0\sigma\omega}} = \sqrt{\frac{2}{(4\pi
\times 10^{-7})(5.96\times 10^{7})(2\pi\times 10^{9})}} = 2.1\,\si{
\micro\meter},
\]
so the current rides in a surface shell two micrometres deep and the
effective resistance of each conductor is set by that shell, not by the
solid cross-section. The surface resistance is $R_s = 1/(\sigma\delta) =
8.0\,\si{\milli\ohm}$ per square, and the resistance per unit length of
the coaxial pair, summed over the inner and outer surfaces, is
\[
R' = \frac{R_s}{2\pi}\left(\frac{1}{a} + \frac{1}{b}\right) =
\frac{8.0\times 10^{-3}}{2\pi}\left(\frac{1}{4.5\times 10^{-4}} +
\frac{1}{1.47\times 10^{-3}}\right) = 3.7\,\si{\ohm\per\meter}.
\]
The conductor contribution to attenuation is $\alpha_c = R'/(2Z_0) =
3.7/(2\times 47) = 0.039\,\si{\per\meter}$, equivalently
\[
\alpha_c\,(\si{\decibel\per\meter}) = 8.686\,\alpha_c = 0.34\,\si{
\decibel\per\meter},
\]
or $34\,\si{\decibel}$ per $100\,\si{\meter}$, in the range a datasheet
quotes for a thin cable of this class at a gigahertz. Because $\delta
\propto f^{-1/2}$, the conductor loss grows as $\sqrt{f}$, the rising
blue curve of \cref{fig:vol04ch08:coax-attenuation}. The dielectric
loss adds $\alpha_d = \tfrac12\omega\sqrt{\mu\varepsilon}\,\tan\delta_d$,
where the loss tangent of good polyethylene is $\tan\delta_d \approx
2\times 10^{-4}$; at $1\,\si{\giga\hertz}$ this contributes only about
$0.027\,\si{\decibel\per\meter}$, well below the conductor term, and
grows linearly with frequency so that it overtakes the conductor loss
only in the tens of gigahertz. The cable is conductor-loss-limited
across the band of interest.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-coax-attenuation.tex}

Over the $30\,\si{\meter}$ run at $1\,\si{\giga\hertz}$ the total loss
is $(0.34 + 0.05)(30) \approx 12\,\si{\decibel}$, which removes
$1 - 10^{-12/10} = 94\,\%$ of the input power before it reaches the far
end. The number is the working content of the cable choice: a thicker,
lower-loss cable would cut it, at higher cost and bend radius, and the
designer trades the loss budget against the installation. A useful rule
falls out of holding $Z_0 = \tfrac{1}{2\pi}\sqrt{\mu/\varepsilon}\ln(b/a)$
fixed while minimising $\alpha_c$: the conductor loss of an air-line
coaxial cable is least at $b/a = 3.6$, which is $Z_0 = 77\,\si{\ohm}$,
the historical reason that low-loss cable-television and antenna feeds
settled near $75\,\si{\ohm}$ rather than the $50\,\si{\ohm}$ chosen for
power handling \cite[ch.~3]{text:pozar-microwave}.

\subsection{Voltage and power ratings}

The cable fails by dielectric breakdown when the radial field at the
inner conductor, where it is strongest, exceeds the dielectric strength.
For a line carrying voltage $V$ between the conductors the peak field is
at $r = a$,
\[
E_{\max} = \frac{V}{a\ln(b/a)},
\]
the result of integrating the $1/r$ field of Section 8.2. Polyethylene
holds about $E_b = 20\,\si{\mega\volt\per\meter}$ before it punctures;
applying a working safety factor of four against the surface
imperfections and voids that lower the practical strength, we cap the
peak field at $5\,\si{\mega\volt\per\meter}$, so the rated voltage is
\[
V_{\text{rated}} = E_{\text{cap}}\,a\ln(b/a) = (5\times 10^{6})(4.5
\times 10^{-4})(1.184) = 2.7\,\si{\kilo\volt}.
\]
The matched-load power the cable can carry at this voltage is
\[
P_{\text{rated}} = \frac{V_{\text{rated}}^{2}}{2 Z_0} = \frac{(2.7
\times 10^{3})^{2}}{2\times 47} = 78\,\si{\kilo\watt}
\]
at the peak of a sinusoid, a continuous rating limited in practice by
the dielectric loss heating the polyethylene rather than by breakdown:
the $12\,\si{\decibel}$ of loss over $30\,\si{\meter}$ becomes heat in
the cable, and the temperature rise of the inner conductor, the hottest
point, falls back to the conduction problem of Volume IV Chapter 5
through the dielectric to the outer braid. The breakdown limit and the
thermal limit cross at a frequency-dependent power, and the lower of the
two governs at any given frequency.

\subsection{The verdict, and a pulsed-power variant}

The specification reads: $47\,\si{\ohm}$ characteristic impedance,
velocity factor $0.67$ giving $150\,\si{\nano\second}$ of delay over the
run, $12\,\si{\decibel}$ of loss at $1\,\si{\giga\hertz}$ over
$30\,\si{\meter}$, a $2.7\,\si{\kilo\volt}$ voltage rating, and a
breakdown-limited matched power near $78\,\si{\kilo\watt}$. Every number
traces to one of the four equations: Gauss's law set the radial field
and the capacitance, Amp\`ere's law set the inductance, the skin-depth
solution of the diffusion equation set the loss, and the Poynting
balance carried the power along the axis between the conductors. The
same cable feeds a pulsed-power load when the source is a charged
capacitor bank rather than a continuous generator, and the analysis
shifts from steady impedance to the transient of the next worked
example, where the cable's own $L'$ and $C'$ become part of the
discharge circuit.

\subsection{A capacitor bank discharged into a load}

A pulsed-power supply stores energy slowly and releases it quickly. A
bank of $C = 200\,\si{\micro\farad}$ charged to $V_0 = 5\,\si{\kilo
\volt}$ holds
\[
U = \tfrac12 C V_0^{2} = \tfrac12(200\times 10^{-6})(5\times 10^{3})^{2}
= 2.5\,\si{\kilo\joule},
\]
the energy of a fast rifle round, delivered in microseconds. Released
through a load resistance $R$ and the unavoidable circuit inductance
$L$, the current obeys the series-$RLC$ equation $L\,\ddot q + R\,\dot q
+ q/C = 0$, whose behaviour is set by the damping ratio $\zeta =
\tfrac{R}{2}\sqrt{C/L}$. For $L = 2\,\si{\micro\henry}$ of bus and lead
inductance and a load $R = 0.5\,\si{\ohm}$,
\[
\zeta = \frac{R}{2}\sqrt{\frac{C}{L}} = \frac{0.5}{2}\sqrt{\frac{200
\times 10^{-6}}{2\times 10^{-6}}} = 2.5,
\]
which is overdamped, $\zeta > 1$, so the current rises and falls without
reversal. The natural frequency is $\omega_0 = 1/\sqrt{LC} = 1/\sqrt{(2
\times 10^{-6})(200\times 10^{-6})} = 5.0\times 10^4\,\si{\per\second}$,
and the peak current is of order $V_0/R = 10\,\si{\kilo\ampere}$ reached
in a few of the $1/\omega_0 = 20\,\si{\micro\second}$ time units. The
critical resistance that delivers the energy fastest without ringing is
\[
R_{\text{crit}} = 2\sqrt{\frac{L}{C}} = 2\sqrt{\frac{2\times 10^{-6}}{
200\times 10^{-6}}} = 0.2\,\si{\ohm},
\]
the boundary of \cref{fig:vol04ch08:capacitor-discharge}. A load below
$R_{\text{crit}}$ rings, reverses polarity, and recharges the bank
backwards, the failure that destroys film and electrolytic banks not
rated for voltage reversal; a load at or above it discharges in a single
clean pulse. The bank designer sizes the series resistance, or adds a
crowbar diode that conducts only on reversal, to keep the operating
point on the safe side of the line.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-capacitor-discharge.tex}

The energy density of the bank measures the cost of pulsed power. A
$200\,\si{\micro\farad}$, $5\,\si{\kilo\volt}$ film capacitor occupies
about $3\,\si{\litre}$, so the stored $2.5\,\si{\kilo\joule}$ gives an
energy density near $0.8\,\si{\joule\per\centi\meter\cubed}$, three orders
of magnitude below a lithium cell of the same volume. Capacitors win on
power density, not energy density: they deliver their stored joules in
microseconds, where a battery would take seconds, and the application
that needs $10\,\si{\kilo\ampere}$ for a microsecond has no other source.
The equivalent series resistance of the bank, a fraction of an ohm,
dissipates $I^{2}\text{ESR}$ during the pulse and sets how often the
bank can be cycled before its own heating forces a pause, the duty-cycle
limit of every pulsed-power machine from a spot welder to a fusion
driver.

\section{A third case study: a brushless DC machine end to end}\pathtag{standard}

The transformer moved field energy across a static magnetic core and the
coaxial line guided it between two conductors; the rotating machine
converts it between electrical and mechanical form, and the design joins
every result of the chapter to a turning shaft. We take a permanent-magnet
brushless DC machine of the size that drives a small electric scooter or a
power tool, rated for $300\,\si{\watt}$ of mechanical output at $3000$
revolutions per minute, and carry it from the air-gap field through the
torque constant, the back-EMF, the winding resistance and copper loss, the
core and friction losses, the efficiency, and the temperature rise, ending
in numbers a machine designer would commit to a drawing. Faraday's law
sets the back-EMF, the Lorentz force sets the torque, Amp\`ere's law and
the reluctance analogy set the air-gap field that the magnets drive across
the gap, and the Poynting balance accounts for every watt that does not
reach the shaft. The single-loop torque result of the worked examples
below is the kernel; this study wraps it into a real machine with magnets,
a slotted stator, and a loss budget \cite[ch.~6]{text:griffiths2017}.

\subsection{Air-gap field from the magnets and the reluctance path}

The rotor carries surface-mounted permanent magnets of remanence
$B_r = 1.2\,\si{\tesla}$ (a sintered neodymium-iron-boron grade) and
radial thickness $\ell_m = 3\,\si{\milli\meter}$, separated from the
stator iron by a mechanical air gap $g = 0.5\,\si{\milli\meter}$. The
magnet is a flux source in series with its own reluctance and the gap
reluctance, the magnetic circuit of the chapter's reluctance box applied
to a moving member. Treating the magnet as a source of magnetomotive force
$\mathcal{F}_m = H_c\ell_m$ driving flux through the series reluctance of
the magnet and the gap, and taking the iron reluctance as negligible
because $\mu_r \gg 1$ in the stator and rotor back-iron, the working
air-gap flux density is
\[
B_g = B_r\,\frac{\ell_m}{\ell_m + \mu_{r,m} g} = (1.2)\,\frac{3.0}{3.0 +
(1.05)(0.5)} = 1.02\,\si{\tesla},
\]
using the recoil permeability $\mu_{r,m}\approx 1.05$ of the magnet. The
gap is short and the magnet thick, so most of the remanence appears across
the gap; a longer gap or a thinner magnet would collapse $B_g$, which is
why the mechanical clearance in such machines is held to a few tenths of a
millimetre. This $B_g$ is the field the conductors sit in, and every
torque and voltage number traces back to it.

\subsection{Torque constant and back-EMF constant}

The stator carries $Z$ active conductors of length $L$ at rotor radius
$R$, each in the air-gap field $B_g$. A conductor carrying current $i$
feels the Lorentz force $F = B_g L i$, and the torque it contributes about
the shaft is $F R$; summing the conductors that are under a pole at any
instant gives a shaft torque proportional to current,
\[
\tau = k_t\,I, \qquad k_t = \frac{Z}{2\pi}\,B_g L R\,\cdot k_w,
\]
where $k_w \approx 0.95$ is a winding factor accounting for the conductors
that lie near a pole edge in a weaker field. Take $Z = 240$ conductors
(an eight-pole, twelve-slot machine wound with $20$ turns per slot),
stack length $L = 40\,\si{\milli\meter}$, and rotor radius $R = 25\,\si{
\milli\meter}$:
\[
k_t = \frac{240}{2\pi}(1.02)(0.040)(0.025)(0.95) = 0.037\,\si{\newton
\meter\per\ampere}.
\]
The back-EMF constant equals the torque constant in SI units, a
consequence of energy conservation that the motional-EMF balance of
Section 8.8 made explicit: the mechanical power $\tau\omega$ must equal
the electrical power $\mathcal{E} I$ in the lossless limit, so with
$\tau = k_t I$ and $\mathcal{E} = k_e\omega$ the equality $\tau\omega =
\mathcal{E}I$ forces $k_t = k_e$. Numerically $k_e = 0.037\,\si{\volt
\second\per\radian}$, so at the rated $\omega = 2\pi(3000)/60 = 314\,\si{
\radian\per\second}$ the back-EMF is
\[
\mathcal{E} = k_e\omega = (0.037)(314) = 11.6\,\si{\volt}
\]
per phase, the voltage the spinning magnets induce in the windings before
any current flows. A drive bus of $24\,\si{\volt}$ leaves headroom above
this back-EMF to push current through the winding resistance, the margin
that sets the top speed.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-bldc-backemf.tex}

The flat-topped back-EMF of \cref{fig:vol04ch08:bldc-backemf} is the
design feature that makes the machine a brushless DC machine rather than a
sinusoidal one. The drive energises the two phases whose back-EMFs sit on
their plateaus during each $60^{\circ}$ electrical step, so the product of
current and back-EMF, and therefore the torque, holds nearly constant
through the electrical cycle. The departure of the real waveform from the
ideal trapezoid is the torque ripple, the quantity a smooth-running drive
works to suppress.

\subsection{Copper loss, core loss, and the rated current}

The rated mechanical output fixes the rated torque: $\tau_{\text{rated}}
= P/\omega = 300/314 = 0.96\,\si{\newton\meter}$, which at $k_t = 0.037\,
\si{\newton\meter\per\ampere}$ needs a phase current of
\[
I_{\text{rated}} = \frac{\tau_{\text{rated}}}{k_t} = \frac{0.96}{0.037}
= 26\,\si{\ampere}.
\]
Two phases conduct at once in the six-step scheme, so the per-conductor
current is set by this value through the energised pair. Sizing the
winding for a current density of $5\,\si{\ampere}$ per square millimetre,
each conductor needs about $5.2\,\si{\milli\meter\squared}$ of copper; the
total conductor length is $Z L = (240)(0.040) = 9.6\,\si{\meter}$ of
active length, and adding end turns roughly doubles it to $L_{\text{cu}}
\approx 19\,\si{\meter}$. At the copper resistivity $1.72\times 10^{-8}\,
\si{\ohm\meter}$ the line-to-line resistance is near $R_{\text{ph}} =
0.063\,\si{\ohm}$ per phase, so the copper loss with two phases conducting
is
\[
P_{\text{cu}} = 2 I_{\text{rated}}^{2} R_{\text{ph}} = 2(26)^{2}(0.063)
= 85\,\si{\watt}
\]
at full torque, the dominant loss of the machine. The core loss is the
hysteresis and eddy-current dissipation in the stator teeth and back-iron,
carrying the same air-gap flux reversed at the electrical frequency
$f_e = (p/2)(n/60) = (8/2)(3000/60) = 200\,\si{\hertz}$ for the eight-pole
rotor. At $B_g \approx 1.0\,\si{\tesla}$ and $200\,\si{\hertz}$ a thin
silicon-steel lamination dissipates about $7\,\si{\watt\per\kilogram}$,
and with a stator iron mass near $0.5\,\si{\kilogram}$ the core loss is
\[
P_{\text{core}} \approx (7)(0.5) = 3.5\,\si{\watt},
\]
small against the copper loss at full torque but, like the transformer's
core loss, fixed independent of load. Friction and windage add a few more
watts; we lump the speed-dependent fixed losses at $P_{\text{fix}}
\approx 20\,\si{\watt}$ at the rated speed, the constant amber line of
\cref{fig:vol04ch08:bldc-efficiency}.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-bldc-efficiency.tex}

\subsection{Efficiency, temperature rise, and the verdict}

The total loss at the rated point is $P_{\text{loss}} = P_{\text{cu}} +
P_{\text{fix}} = 85 + 20 = 105\,\si{\watt}$, so the efficiency at full
torque is
\[
\eta = \frac{P_{\text{out}}}{P_{\text{out}} + P_{\text{loss}}} =
\frac{300}{300 + 105} = 0.74,
\]
modest because this small machine is worked hard at its rated current. The
efficiency peaks not at full torque but where the copper loss equals the
fixed loss, at $\tau = \sqrt{P_{\text{fix}}/(P_{\text{cu}}/\tau_{\text{
rated}}^{2})} = \tau_{\text{rated}}\sqrt{P_{\text{fix}}/P_{\text{cu}}} =
(0.96)\sqrt{20/85} = 0.47\,\si{\newton\meter}$, about half rated torque,
the crossover of \cref{fig:vol04ch08:bldc-efficiency}. A machine that
spends its life at half load should be wound so this peak sits at the load
it actually carries, the same lesson the transformer taught. The
$85\,\si{\watt}$ of copper loss is dumped into the stator windings, whose
exposed surface of about $0.02\,\si{\meter\squared}$ removes heat by
conduction to the housing and convection from it; at a combined
coefficient near $40\,\si{\watt\per\meter\squared}$ per kelvin (the higher
value reflecting forced air over a fan-cooled motor, from the convection
results of Chapter 5) the steady winding temperature rise is
\[
\Delta T = \frac{P_{\text{loss}}}{h A_s} = \frac{105}{(40)(0.02)} =
131\,\si{\kelvin},
\]
which over a $40\,\degC$ ambient reaches a winding temperature near
$171\,\degC$, above the $155\,\degC$ rating of class-F insulation. The
verdict is a machine that meets its torque and speed but runs too hot at
continuous full load; the fix is to derate the continuous rating to about
seventy percent of peak torque, where the copper loss falls to half and
the rise drops under the insulation limit, while the peak torque remains
available for the short bursts of acceleration a scooter or a tool
demands. Every number traces to one of the four equations: the magnet
flux and the reluctance path set $B_g$, the Lorentz force set the torque
constant, Faraday's law set the equal back-EMF constant, and the resistive
$\vect{J}\cdot\vect{E}$ of the Poynting balance set the copper loss and the
heat the temperature rise must remove. The same procedure sizes a
generator running the conversion backwards, with the shaft driven and the
windings delivering power, where the back-EMF becomes the source and the
sign of the torque reverses.

\begin{mastery}
The machine has a hidden torque the slotted geometry produces with no
current at all. As the rotor turns, the magnets prefer angular positions
that minimise the reluctance of their flux path, aligning a magnet pole
with a stator tooth rather than a slot opening, and the energy
$U = \tfrac12\Phi^{2}\mathcal{R}(\theta)$ varies with rotor angle through
the angle-dependent reluctance $\mathcal{R}(\theta)$. The resulting
\engterm{cogging torque} is $\tau_{\text{cog}} = -\dd U/\dd\theta =
\tfrac12\Phi^{2}\,\dd\mathcal{R}/\dd\theta$, present even when the windings
are open, felt as the notchy detenting when a brushless motor is turned by
hand. Its fundamental period is set by the least common multiple of the
pole count and the slot count, so the twelve-slot eight-pole machine
cogs at $\operatorname{lcm}(12,8) = 24$ cycles per revolution, a high
frequency of small amplitude rather than a low frequency of large
amplitude. Designers suppress it by skewing the slots or the magnets by
one slot pitch, so the reluctance variation averages along the stack and
$\dd\mathcal{R}/\dd\theta$ nearly cancels, and by shaping the magnet edges
so the flux enters the gap gradually rather than stepping at the pole
boundary. Cogging is the reluctance analogy of the chapter producing a
force directly, with no winding current, purely from the geometry of the
flux path seeking its lowest-reluctance seat.
\end{mastery}

\section{More worked examples: force, torque, and loss}\pathtag{standard}

The chapter's force results deserve cases of their own, since the force
an electromagnetic device exerts is the quantity a customer pays for and
the quantity most often mis-estimated by the factor of one half that
recurs through the chapter.

\subsection{The far field of a bounded source: the multipole ladder}

Gauss's law fixes the total outward flux of a bounded charge distribution
by the net charge alone, and that single number sets the leading term of
the field far away. A distribution confined to a region of size $d$,
viewed from a distance $r \gg d$, produces a field that organises itself
into a ladder of terms falling off ever faster with distance, the
\engterm{multipole expansion}. The potential of a bounded charge cloud is
\[
\phi(r) = \frac{1}{4\pi\varepsilon_0}\left[\frac{Q}{r} +
\frac{\vect{p}\cdot\unitvec{r}}{r^{2}} + \frac{1}{2}
\frac{\sum_{ij}Q_{ij}\hat r_i\hat r_j}{r^{3}} + \cdots\right],
\]
with $Q = \int\rho\,\dd V$ the net charge (monopole), $\vect{p} =
\int\vect{r}'\rho\,\dd V$ the dipole moment, and $Q_{ij}$ the quadrupole
tensor. The field is one power of $r$ steeper than the potential in each
term, so the monopole field falls as $r^{-2}$, the dipole as $r^{-3}$, and
the quadrupole as $r^{-4}$, the ladder of \cref{fig:vol04ch08:dipole-multipole}.
The lowest surviving term dominates far away: a charged object looks like a
point charge, a neutral polar object like a dipole, and a neutral,
dipole-free object like a quadrupole. The engineering payoff is in stray
fields. A single unbalanced conductor presents a monopole of current, its
field dying only as $r^{-1}$ in the magnetic case; a balanced go-and-return
pair presents no net current and its leading field is the dipole of the
loop between them, dying as $r^{-3}$; and a tightly twisted pair reverses
the loop sense every twist, pushing the leading term to a quadrupole that
dies as $r^{-4}$. Each rung of the ladder that a layout climbs is one power
of distance faster that its interference dies, which is the field-theory
reason a twisted, balanced cable is quiet and a single wire is loud.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-dipole-multipole.tex}

\subsection{Force between parallel conductors: the ampere and the busbar}

Two long parallel wires a distance $d$ apart, carrying currents $I_1$ and
$I_2$, exert a force on each other through the field one sets up at the
other. The first wire produces at the second a field $B_1 = \mu_0 I_1/(2\pi
d)$, and the second wire, carrying $I_2$ in that field, feels a force per
unit length
\[
\frac{F}{\ell} = B_1 I_2 = \frac{\mu_0 I_1 I_2}{2\pi d},
\]
attractive when the currents run the same way and repulsive when they
oppose. The relation is the historical definition of the ampere, fixed in
the pre-2019 SI as the current that produces $2\times 10^{-7}\,\si{\newton}$
per metre between wires a metre apart, a definition the 2019 redefinition
replaced by fixing the elementary charge but that still sets the working
size of the force. Two bench wires at $1\,\si{\ampere}$ and $1\,\si{\centi
\meter}$ apart pull on each other with only $2\times 10^{-5}\,\si{\newton}$
per metre, a force too small to feel, which is why ordinary wiring shows no
mechanical effect. The force turns violent under fault current. In a
switchboard a short circuit can drive $50\,\si{\kilo\ampere}$ through
busbars spaced $5\,\si{\centi\meter}$ apart, giving
\[
\frac{F}{\ell} = \frac{(4\pi\times 10^{-7})(5\times 10^{4})^{2}}{2\pi
(0.05)} = 1.0\times 10^{4}\,\si{\newton\per\meter},
\]
a tonne of force per metre trying to tear the bars apart, which is why
switchgear busbars are braced at close intervals and rated for a
mechanical short-circuit withstand as well as a thermal one. The force
scales as the square of the current, so it is the fault, not the load,
that sizes the bracing.

\subsection{Quality factor and ring-down of a resonant tank}

A capacitor and an inductor exchange energy back and forth, and the rate at
which the exchange decays measures how sharply the pair resonates. A series
$RLC$ circuit oscillates at $\omega_0 = 1/\sqrt{LC}$ and loses a fraction of
its stored energy each cycle to the resistance, and the \engterm{quality
factor}
\[
Q = \frac{\omega_0 L}{R} = \frac{1}{R}\sqrt{\frac{L}{C}} = 2\pi
\frac{\text{energy stored}}{\text{energy lost per cycle}}
\]
counts how many radians of oscillation pass before the energy falls by a
factor $e$. A tank with $L = 100\,\si{\micro\henry}$, $C = 100\,\si{\pico
\farad}$, and coil resistance $R = 5\,\si{\ohm}$ resonates at $\omega_0 =
1/\sqrt{(10^{-4})(10^{-10})} = 10^{7}\,\si{\per\second}$, a frequency
$f_0 = 1.6\,\si{\mega\hertz}$, with
\[
Q = \frac{(10^{7})(10^{-4})}{5} = 200.
\]
An excited tank rings down as $e^{-\omega_0 t/(2Q)}$, so its amplitude
falls to a fraction $1/e$ after $Q/\pi \approx 64$ cycles, and its
resonance is sharp: the bandwidth over which it responds is $\Delta f =
f_0/Q = 8\,\si{\kilo\hertz}$, narrow enough to pick one radio station from
its neighbours. The quality factor is the figure that decides whether a
loosely coupled link transfers power, whether a filter passes a clean band,
and whether an oscillator holds a stable frequency, and the same $Q$
reappears in the resonant tuning of the wireless-charging vignette below,
where a high $Q$ buys the transfer that a weak coupling alone would lose.

\subsection{Magnetic pressure and the pinch}

A magnetic field pushes on the currents that carry it, and the push has the
dimensions of a pressure. The energy density $B^{2}/(2\mu_0)$ of a field is
also the pressure it exerts normal to itself, so a field $B$ on one side of
a current sheet and zero on the other presses on the sheet with
\[
p_{\text{mag}} = \frac{B^{2}}{2\mu_0}.
\]
At $B = 1\,\si{\tesla}$ this is $p = 1/(2\times 4\pi\times 10^{-7}) = 4.0
\times 10^{5}\,\si{\pascal}$, about four atmospheres; at $B = 10\,\si{
\tesla}$ it rises a hundredfold to four hundred atmospheres, the pressure
that tries to burst the windings of a high-field magnet and the reason such
magnets are built with the winding stress, not the field, as the governing
limit. The same pressure squeezes a current-carrying plasma column inward,
the \engterm{pinch}: a column carrying axial current $I$ generates an
azimuthal field $B_\theta = \mu_0 I/(2\pi a)$ at its radius $a$, and the
inward magnetic pressure balances the plasma's kinetic pressure $p = n k_B
T$ when
\[
\frac{B_\theta^{2}}{2\mu_0} = n k_B T,
\]
the Bennett relation that sets the confining current a pinch device needs.
A field of a few tesla confines a plasma at a pressure of atmospheres, far
below what a solid wall could hold at the same temperature, which is why
magnetic confinement is the route to controlled fusion: the field, not a
material wall, carries the pressure. The magnetic pressure is the
chapter's energy density wearing its mechanical face, the same $B^{2}/
(2\mu_0)$ that pulled the solenoid actuator's plunger, here pushing on a
column of moving charge.

\subsection{The conductor-dielectric crossover frequency}

Whether a material behaves as a conductor or a dielectric is not fixed by
the material alone but by the frequency it is asked to work at, and the
crossover is a single number. The Amp\`ere-Maxwell law carries a conduction
current $\sigma\vect{E}$ and a displacement current $\varepsilon\,\partial
\vect{E}/\partial t$, whose ratio for a sinusoid at angular frequency
$\omega$ is the loss tangent $\sigma/(\omega\varepsilon)$. The two currents
are equal at the crossover frequency
\[
f_{\times} = \frac{\sigma}{2\pi\varepsilon},
\]
above which displacement current dominates and the material passes waves as
a dielectric, below which conduction current dominates and it behaves as a
conductor. For seawater, $\sigma \approx 4\,\si{\siemens\per\meter}$ and
$\varepsilon_r \approx 80$, so
\[
f_{\times} = \frac{4}{2\pi(80)(8.85\times 10^{-12})} = 9.0\times 10^{8}\,
\si{\hertz},
\]
nearly a gigahertz: seawater conducts at the kilohertz frequency a
submarine communication link uses, which is why such links reach a
submerged boat, and it turns dielectric only in the microwave band, where a
radar wave reflects off the surface instead of penetrating. The same number
for copper, with $\sigma = 6\times 10^{7}\,\si{\siemens\per\meter}$ and
$\varepsilon_r \approx 1$, sits at $10^{18}\,\si{\hertz}$, in the X-ray
band, so copper is a conductor at every frequency an engineer drives. The
crossover frequency is the reciprocal of the charge-relaxation time up to
the factor $2\pi$, the same $\varepsilon/\sigma$ that set how fast a
conductor neutralises, read now as the frequency at which it stops acting
like one.

\subsection{Thermal noise: the field a warm resistor radiates into a wire}

Every resistor at a finite temperature sets a floor under how small a signal
a circuit can measure, because the thermal motion of its charge carriers
drives a fluctuating voltage across its terminals with no signal applied.
The mean-square open-circuit voltage of a resistance $R$ at temperature $T$,
measured over a bandwidth $\Delta f$, is the Johnson-Nyquist result
\[
\langle V^{2}\rangle = 4 k_B T R\,\Delta f,
\]
independent of the material and the construction of the resistor, a
consequence of the same equipartition that governs a gas. A $1\,\si{\mega
\ohm}$ resistor at room temperature ($T = 300\,\si{\kelvin}$) measured over
a $10\,\si{\kilo\hertz}$ bandwidth gives
\[
\sqrt{\langle V^{2}\rangle} = \sqrt{4(1.38\times 10^{-23})(300)(10^{6})
(10^{4})} = 1.3\times 10^{-5}\,\si{\volt} = 13\,\si{\micro\volt},
\]
a floor that no amplifier can beat, since the amplifier sees the noise
along with the signal. The noise falls with the square root of resistance,
of temperature, and of bandwidth, which is why a sensitive front end uses
the lowest source resistance the design allows, why a radio telescope cools
its first amplifier to a few kelvin, and why narrowing the bandwidth to the
signal's own width is the cheapest noise reduction available. The thermal
noise is the electromagnetic floor of every measurement in the book: the
calibration chapter of Volume I bounded a measurement by its systematic
error, and this is the matching bound from the random side, set by the
temperature of the very resistor that reads the signal.

\subsection{The betatron condition: a two-to-one rule from Faraday's law}

The induced electric field of the changing-flux example accelerates a real
particle in the betatron, and holding that particle on a fixed circle while
it accelerates forces a precise relation between the guiding field and the
average field the flux provides. An electron circulating at radius $R$ in a
guide field $B_g$ has momentum $p = e B_g R$ from the orbit relation, and
Faraday's law accelerates it: the EMF around the orbit is $-\dd\Phi/\dd t$,
so the momentum rises at $\dd p/\dd t = e\mathcal{E}/(2\pi R) = -(e/2\pi R)
\,\dd\Phi/\dd t$. To keep $R$ fixed, $\dd p/\dd t = e R\,\dd B_g/\dd t$, and
equating the two rates gives
\[
2\pi R^{2}\dv{B_g}{t} = \dv{\Phi}{t}, \qquad \Phi = 2\pi R^{2} B_g = 2\bigl(
\pi R^{2}\bigr)B_g,
\]
the \engterm{betatron condition}: the flux through the orbit must be twice
what a uniform field equal to the guide field would give, so the average
field inside the orbit must be exactly twice the field at the orbit. A
betatron therefore shapes its pole faces to pile extra flux through the
central region while holding the edge field at the value the orbit radius
needs, a purely geometric requirement that falls straight out of Faraday's
law combined with the orbit relation. The rule is the cleanest illustration
that the induced field of a changing flux does mechanical work on charge,
turn after turn, with no electrode and no potential drop anywhere in the
loop.

\subsection{Torque on a current loop: the motor}

A rectangular loop of $N$ turns, sides $w$ and $h$, carrying current
$I$, sits in a uniform field $B$ with its plane parallel to the field.
Each of the two sides of length $h$ that run perpendicular to $B$ feels
a force $F = N B I h$, and the two forces form a couple with moment arm
$w$, giving a torque
\[
\tau = N B I h w = N B I A,
\]
with $A = hw$ the loop area. The product $m = N I A$ is the loop's
\engterm{magnetic dipole moment}, so $\tau = m B$ at this orientation
and $\tau = m B\sin\theta$ in general, where $\theta$ is the angle
between the loop's normal and the field. A motor armature of $N = 200$
turns, area $A = 40\,\si{\centi\meter\squared} = 4.0\times 10^{-3}\,\si{
\meter\squared}$, carrying $I = 3\,\si{\ampere}$ in a field $B = 0.4\,
\si{\tesla}$ produces a peak torque
\[
\tau = (200)(0.4)(3)(4.0\times 10^{-3}) = 0.96\,\si{\newton\meter}.
\]
At a shaft speed of $3000$ revolutions per minute, $\omega = 2\pi\times
3000/60 = 314\,\si{\per\second}$, the mechanical power at this torque is
$P = \tau\omega = (0.96)(314) = 300\,\si{\watt}$, a small fractional-
horsepower motor. The torque falls as $\sin\theta$ through the rotation,
which is why a practical motor uses many armature coils at staggered
angles, or a commutator that switches the current to keep the
energised coil near the peak-torque orientation, so the shaft sees a
steadier torque than any single loop could supply.

\subsection{Pull force of a solenoid actuator}

A solenoid actuator pulls a ferromagnetic plunger into a coil. The force
is the gradient of the magnetic energy with respect to the plunger
position, and at fixed magnetomotive force $\mathcal{F} = NI$ across a
working air gap $g$ the energy stored in the gap is $U = \tfrac12
\mathcal{F}^{2}/\mathcal{R}_g$ with gap reluctance $\mathcal{R}_g = g/
(\mu_0 A)$, so $U = \tfrac12\mu_0 A\mathcal{F}^{2}/g$ and the pull force
is
\[
F = -\dv{U}{g} = \frac{\mu_0 A (NI)^{2}}{2 g^{2}} = \frac{B^{2} A}{2
\mu_0},
\]
writing $B = \mu_0 NI/g$ for the gap field. The form $B^2 A/(2\mu_0)$ is
the magnetic pressure $B^2/(2\mu_0)$ acting over the pole area, the
magnetic counterpart of the electrostatic pressure of Section 8.8. A
plunger of pole area $A = 2\,\si{\centi\meter\squared} = 2.0\times
10^{-4}\,\si{\meter\squared}$ driven to a gap field $B = 1.0\,\si{\tesla}
$ pulls with
\[
F = \frac{(1.0)^{2}(2.0\times 10^{-4})}{2(4\pi\times 10^{-7})} = 80\,
\si{\newton},
\]
enough to throw a heavy relay or a hydraulic valve. The force scales as
$1/g^2$ at fixed ampere-turns, the steep curve of
\cref{fig:vol04ch08:actuator-force}: the actuator is weakest at the open
gap where its stroke begins and strongest as it seats, the opposite of
what a constant-force load wants, which is the design tension every
solenoid actuator must resolve with a shaped pole face or a graded
return spring.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-actuator-force.tex}

\subsection{Splitting the core loss: hysteresis against eddy current}

The transformer case study lumped the core loss into a single figure of
$1.1\,\si{\watt\per\kilogram}$. Separating it into its two mechanisms is
the measurement that tells a core designer which lever to pull. Total
specific core loss at peak flux density $B_{\text{pk}}$ and frequency
$f$ follows the Steinmetz decomposition
\[
P_{\text{core}} = k_h f B_{\text{pk}}^{n} + k_e f^{2} B_{\text{pk}}^{2},
\]
where the first term is hysteresis, linear in frequency because the
loop is traversed once per cycle, and the second is eddy current,
quadratic in frequency because the induced eddy EMF and hence the eddy
current both rise with $f$. The Steinmetz exponent $n$ sits near $1.6$
to $2.0$ for silicon steel. Dividing through by $f$,
\[
\frac{P_{\text{core}}}{f} = k_h B_{\text{pk}}^{n} + k_e f
B_{\text{pk}}^{2},
\]
is linear in $f$ at fixed $B_{\text{pk}}$, so loss measured at two
frequencies separates the two terms: the intercept is hysteresis, the
slope is eddy current. For a steel measured at $1.0\,\si{\tesla}$ giving
$0.6\,\si{\watt\per\kilogram}$ at $50\,\si{\hertz}$ and $1.6\,\si{\watt
\per\kilogram}$ at $100\,\si{\hertz}$, the points $(50, 0.012)$ and
$(100, 0.016)$ in $(f, P/f)$ coordinates give a slope of $8.0\times
10^{-5}$ and an intercept of $0.008$, so at $50\,\si{\hertz}$ the
hysteresis share is $0.008\times 50 = 0.4\,\si{\watt\per\kilogram}$ and
the eddy share is $0.2\,\si{\watt\per\kilogram}$, a two-to-one split.
The lever each term responds to differs: hysteresis falls with a
cleaner, more grain-oriented steel that narrows the B-H loop, while eddy
current falls with thinner laminations that shrink the loops the induced
currents can close, since $k_e \propto t^{2}$ in the lamination
thickness $t$. A designer who measures a loss dominated by eddy current
orders thinner laminations; one who measures hysteresis-dominated loss
orders a better grade of steel. The single lumped number hides this
choice; the two-frequency split exposes it.

The hysteresis term traces directly to the area of the $B$--$H$ loop of
\cref{fig:vol04ch08:hysteresis-loop}. Driving the core once around the loop
does work equal to $\oint H\,\dd B$ per unit volume, the enclosed area, and
doing it $f$ times a second dissipates that area times $f$ as heat, which is
why the hysteresis loss is linear in frequency and set by the loop's width
rather than by how fast it is traversed. A soft steel narrows the loop
toward a single-valued curve and collapses the area; a permanent-magnet
grade fattens it on purpose, storing the wide loop's remanence $B_r$ and
coercivity $H_c$ as the properties that hold a field, and paying the large
area only because such a magnet is magnetised once and not cycled. The
Steinmetz exponent $n$ near $1.6$ to $2.0$ is the empirical statement that
the loop area grows a little faster than linearly with peak flux density,
because the tips of the loop widen as the material is driven harder toward
its saturation. The lever the two-frequency split identifies is now visible
on the curve: a designer fighting hysteresis loss works to narrow the loop,
while one fighting eddy loss works to shrink the conducting paths the
induced currents close, two independent changes to two independent parts of
the same measured number.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-hysteresis-loop.tex}

\begin{mastery}
The skin effect that governed the cable's loss also caps the lamination
thickness that the eddy-current split rewards. A lamination thicker than
about two skin depths at the operating frequency gains nothing from
being thicker, because the flux cannot penetrate the interior: the
induced eddy currents shield the core's centre exactly as they shield a
conductor's interior at radio frequency. At $50\,\si{\hertz}$ the skin
depth in silicon steel, with its high permeability $\mu_r\sim 5000$ and
conductivity $\sigma\sim 2\times 10^{6}\,\si{\siemens\per\meter}$, is
\[
\delta = \sqrt{\frac{2}{\mu\sigma\omega}} = \sqrt{\frac{2}{(5000)(4\pi
\times 10^{-7})(2\times 10^{6})(2\pi\times 50)}} \approx 0.71\,\si{\milli
\meter},
\]
so a lamination thicker than about $1.4\,\si{\milli\meter}$ carries flux
only in its outer $0.7\,\si{\milli\meter}$ and wastes the rest. Standard
mains-frequency laminations are $0.35$ to $0.5\,\si{\milli\meter}$ for
exactly this reason, comfortably under the skin-depth ceiling, and a
transformer designed to run at $400\,\si{\hertz}$ in an aircraft, where
$\delta$ is smaller by $\sqrt{8}$, uses thinner laminations still. The
same physics that forced thin plating on the radio-frequency cable
forces thin laminations on the power-frequency core: at every frequency
there is a thickness beyond which the conductor or the core stops
earning its volume.
\end{mastery}

\subsection{The Hall voltage: measuring a field with the Lorentz force}

The Lorentz force that drives a motor also measures a field. A flat
conductor of thickness $t$ carries current $I$ along its length while
sitting in a transverse field $B$ normal to its face. Each charge carrier
drifting at velocity $v_d$ feels a sideways force $q v_d B$ that pushes it
to one edge, and charge accumulates there until the transverse electric
field it builds balances the magnetic force, $q E_H = q v_d B$. The
resulting \engterm{Hall voltage} across the width $w$ is $V_H = E_H w =
v_d B w$, and writing the drift velocity in terms of the current through
the carrier density $n$, $I = n q v_d (w t)$, gives
\[
V_H = \frac{I B}{n q t},
\]
linear in the field $B$, which is what makes the device a field sensor.
For a thin semiconductor Hall plate with carrier density $n = 10^{22}\,
\si{\per\meter\cubed}$, thickness $t = 50\,\si{\micro\meter}$, carrying
$I = 5\,\si{\milli\ampere}$ in a field $B = 0.1\,\si{\tesla}$,
\[
V_H = \frac{(5\times 10^{-3})(0.1)}{(10^{22})(1.6\times 10^{-19})
(50\times 10^{-6})} = 6.3\times 10^{-3}\,\si{\volt} = 6.3\,\si{\milli
\volt}.
\]
A semiconductor wins over a metal because its carrier density is some
seven orders of magnitude lower than copper's $8.5\times 10^{28}\,\si{
\per\meter\cubed}$, so the same current and field produce a Hall voltage
seven orders larger: a metal plate would give nanovolts, a semiconductor
gives millivolts. The Hall sensor that calibrated the chapter's project,
and the rotor-position sensor that tells a brushless drive when to
commutate, are this equation in a package.

\subsection{Eddy-current braking on a moving conductor}

A conductor moving through a non-uniform field is braked by the currents
its own motion induces, the mechanism of a contactless brake and of the
damping in a balance scale. Consider a conducting sheet of thickness $d$
and conductivity $\sigma$ moving at speed $v$ through a region of field
$B$ confined to a pole of face area $A$. The motional EMF $\mathcal{E} =
B L v$ across the pole drives eddy currents in the sheet against its own
resistance, and the dissipated power is the retarding force times the
speed. A working estimate takes the induced current density as
$J \sim \sigma v B$ over the swept volume $A d$, so the dissipated power
is
\[
P \sim \sigma (vB)^{2}\,A d,
\]
and the retarding force is $F = P/v = \sigma v B^{2} A d$, proportional to
speed. For an aluminium sheet ($\sigma = 3.5\times 10^{7}\,\si{\siemens
\per\meter}$), thickness $d = 2\,\si{\milli\meter}$, moving at $v = 5\,\si{
\meter\per\second}$ through a pole of face $A = 4\,\si{\centi\meter
\squared} = 4\times 10^{-4}\,\si{\meter\squared}$ at $B = 0.5\,\si{
\tesla}$,
\[
F = (3.5\times 10^{7})(5)(0.5)^{2}(4\times 10^{-4})(2\times 10^{-3})
= 35\,\si{\newton},
\]
a force that opposes the motion and grows with speed. The brake is
strongest when moving fastest and vanishes at rest, the opposite of a
friction brake, which is why eddy-current brakes hold a steady descent
speed but cannot clamp a stopped load. The force scales with conductivity
and field squared, so a copper sheet brakes harder than aluminium and a
stronger magnet brakes much harder; the energy removed appears as heat in
the sheet, the same $\vect{J}\cdot\vect{E}$ dissipation that heated the
transformer core.

\subsection{Mutual inductance and the coupling coefficient}

Two coils share flux, and the fraction they share decides whether a
transformer works or an interference source couples. A long primary
solenoid of $n_1$ turns per metre, cross-section $A$, carrying current
$I_1$ produces an interior field $B = \mu_0 n_1 I_1$, and a secondary of
$N_2$ turns wound over a length $\ell$ of the same core links flux
$N_2 B A$. The mutual inductance is the linked flux per unit primary
current,
\[
M = \frac{N_2 B A}{I_1} = \mu_0 n_1 N_2 A = \mu_0 n_1 (n_2\ell) A,
\]
writing $N_2 = n_2\ell$. For $n_1 = n_2 = 2000\,\si{\per\meter}$, overlap
length $\ell = 0.1\,\si{\meter}$, and cross-section $A = 4\,\si{\centi
\meter\squared} = 4\times 10^{-4}\,\si{\meter\squared}$,
\[
M = (4\pi\times 10^{-7})(2000)(2000\times 0.1)(4\times 10^{-4}) = 2.0
\times 10^{-4}\,\si{\henry} = 0.20\,\si{\milli\henry}.
\]
The coupling coefficient $k = M/\sqrt{L_1 L_2}$ measures how completely
the two coils share flux, bounded by $0 \le k \le 1$. With self
inductances $L_1 = \mu_0 n_1^{2} A \ell_1$ and $L_2 = \mu_0 n_2^{2} A
\ell$ for primary length $\ell_1 = 0.2\,\si{\meter}$, the values are
$L_1 = 0.40\,\si{\milli\henry}$ and $L_2 = 0.20\,\si{\milli\henry}$,
giving $k = (0.20)/\sqrt{(0.40)(0.20)} = 0.71$. A coupling of $0.71$ means
half the flux links both coils and half leaks; a power transformer drives
$k$ above $0.99$ with an interleaved winding on a closed core, while a
loosely coupled antenna or a stray interference path may sit at $k <
0.01$. The leakage that lowers $k$ is the same leakage flux the no-monopole
section named as the designer's adversary, now given a number.

\subsection{Reflection of a wave at a dielectric interface}

The boundary conditions on $\vect{E}$ and $\vect{H}$ decide how much of a
wave crosses an interface and how much returns, the calculation behind
every anti-reflection coating and every impedance match. A plane wave in
medium 1 of intrinsic impedance $\eta_1 = \sqrt{\mu_1/\varepsilon_1}$
strikes a flat boundary with medium 2 of impedance $\eta_2$ at normal
incidence. Continuity of tangential $\vect{E}$ and tangential $\vect{H}$
across the boundary, the interface conditions of Section 8.7, force the
amplitude reflection coefficient
\[
\Gamma = \frac{\eta_2 - \eta_1}{\eta_2 + \eta_1},
\]
the optical and electromagnetic twin of the transmission-line mismatch of
the coaxial study. For a wave in air ($\eta_1 = 377\,\si{\ohm}$, the
impedance of free space) striking glass of $\varepsilon_r = 2.25$, the
glass impedance is $\eta_2 = 377/\sqrt{2.25} = 251\,\si{\ohm}$, so
\[
\Gamma = \frac{251 - 377}{251 + 377} = -0.20,
\]
and the reflected power fraction is $|\Gamma|^{2} = 0.040$, the four
percent that every uncoated glass surface returns, the glare on a window
and the loss at each face of a lens. The negative sign records the
half-cycle phase flip on reflection from the slower medium, the same flip
the overdamped capacitor bank avoided and the standing-wave pattern of
Chapter 10 depends on. A quarter-wave coating of impedance
$\eta_c = \sqrt{\eta_1\eta_2} = \sqrt{(377)(251)} = 308\,\si{\ohm}$
cancels the reflection entirely at its design wavelength, the
geometric-mean match that the same boundary conditions predict and that
every camera lens carries.

\subsection{Charged particle in a magnetic field: the mass spectrometer}

A particle of charge $q$ and mass $m$ moving with speed $v$ perpendicular
to a uniform field $B$ feels the Lorentz force $q\vect{v}\times\vect{B}$,
always perpendicular to the velocity, which curves the path into a circle
without changing the speed. Setting the magnetic force equal to the
centripetal requirement $mv^{2}/r$ gives $qvB = mv^{2}/r$, so the radius
is
\[
r = \frac{mv}{qB},
\]
proportional to momentum and inversely proportional to the field. The
angular frequency of the orbit, the \engterm{cyclotron frequency}, is
\[
\omega_c = \frac{v}{r} = \frac{qB}{m},
\]
independent of speed and radius: every particle of a given
charge-to-mass ratio circulates at the same frequency whatever its
energy, the property that lets a cyclotron accelerate a particle through
many turns with a fixed-frequency drive. A singly ionised molecule of
mass $100\,\text{u}$ ($1.66\times 10^{-25}\,\si{\kilogram}$) accelerated
through $1\,\si{\kilo\volt}$ reaches a speed from $\tfrac12 mv^{2} = qV$
of $v = \sqrt{2qV/m} = \sqrt{2(1.6\times 10^{-19})(10^{3})/(1.66\times
10^{-25})} = 4.4\times 10^{4}\,\si{\meter\per\second}$, and in a field
$B = 0.5\,\si{\tesla}$ its orbit radius is
\[
r = \frac{mv}{qB} = \frac{(1.66\times 10^{-25})(4.4\times 10^{4})}{
(1.6\times 10^{-19})(0.5)} = 0.091\,\si{\meter},
\]
about nine centimetres. A molecule of twice the mass at the same
accelerating voltage moves slower by $\sqrt{2}$ but carries $\sqrt{2}$
times the momentum, so its radius is larger by $\sqrt{2}$, and the two
strike a detector at separated positions. This is the working principle
of the magnetic-sector mass spectrometer: the field sorts ions by their
mass-to-charge ratio along a position the radius equation predicts
exactly. The same equation, read the other way, is the design constraint
on every storage ring and beam line, where the field must rise in lockstep
with the beam momentum to hold the orbit radius fixed.

The radius line of \cref{fig:vol04ch08:cyclotron-orbit} makes the sorting
quantitative. Because $r = p/(qB)$ is linear in momentum, a beam of fixed
momentum entering a uniform-field region fans out along the radius axis by
charge alone, and a beam of mixed masses accelerated through one voltage
fans out by mass, since two singly charged ions dropped through the same
potential carry momenta in the ratio of the square roots of their masses.
The mass resolution the instrument reaches is set by how tightly the entering
beam is collimated against how far the radius line separates neighbouring
masses, so a sector magnet resolving one part in a thousand needs a beam
whose angular and energy spread stay under that fraction. The cyclotron
frequency $\omega_c = qB/m$ carries the dual lesson: it is independent of
radius, so a particle spiralling out as it gains energy keeps step with a
fixed-frequency accelerating voltage, until the energy climbs high enough
that the relativistic mass rise detunes it, the limit that forced the
synchrocyclotron and then the synchrotron to sweep either the frequency or
the field as the beam accelerates.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-cyclotron-orbit.tex}

\subsection{The induced electric field inside a changing-flux region}

Faraday's law in integral form gives the magnitude of the induced
electric field directly when the geometry is symmetric, with no need for
the potentials. A uniform magnetic field $B(t)$ fills a long cylindrical
region of radius $R$, pointing along the axis and rising at a steady rate
$\dd B/\dd t$. By the axial symmetry the induced electric field circles
the axis with constant magnitude on any circle of radius $r$ centred on
it. Apply $\oint\vect{E}\cdot\dd\vect{\ell} = -\dd\Phi_B/\dd t$ to such a
circle. Inside the field region, $r < R$, the enclosed flux is
$\Phi_B = B\pi r^{2}$ and the circulation is $E(2\pi r)$, so
\[
E(2\pi r) = -\pi r^{2}\dv{B}{t}, \qquad E = -\frac{r}{2}\dv{B}{t},
\]
rising linearly with radius from zero on the axis. Outside the field
region, $r > R$, the enclosed flux is fixed at $B\pi R^{2}$ regardless of
$r$, so
\[
E(2\pi r) = -\pi R^{2}\dv{B}{t}, \qquad E = -\frac{R^{2}}{2r}\dv{B}{t},
\]
falling as $1/r$. For $R = 5\,\si{\centi\meter}$ and a field ramping at
$\dd B/\dd t = 100\,\si{\tesla\per\second}$, the induced field peaks at
the boundary $r = R$ with magnitude $E = (0.05/2)(100) = 2.5\,\si{\volt
\per\meter}$. A changing magnetic field produces a real, circulating
electric field that exists in the empty space around the region, with no
charge anywhere to source it, the cleanest demonstration that the
electric field of Faraday's law is not the gradient of a potential: its
line integral around a closed loop is not zero. The betatron accelerates
electrons by exactly this induced field, ramping the flux through a ring
so the circulating $\vect{E}$ drives the electrons forward turn after
turn, and the eddy-current and transformer losses of the chapter are the
same induced field driving currents in conductors that happen to lie in
the changing flux.

\subsection{Pull-in: the electrostatic instability of a movable plate}

A movable capacitor plate suspended on a spring against an electrostatic
force has a stable position only up to a critical voltage, beyond which it
snaps shut, the instability that governs every electrostatic
microactuator and the cause of stuck microelectromechanical switches. Let
one plate of area $A$ be held by a spring of stiffness $\kappa$ at a
zero-voltage gap $g_0$, and let the gap close to $g$ when voltage $V$ is
applied. The electrostatic attraction from the worked result above is
$F_e = \varepsilon_0 A V^{2}/(2g^{2})$, pulling the plates together, and
the spring pushes back with $F_s = \kappa(g_0 - g)$. Equilibrium balances
them,
\[
\frac{\varepsilon_0 A V^{2}}{2g^{2}} = \kappa(g_0 - g).
\]
The electrostatic force grows as $1/g^{2}$ as the gap closes while the
spring force grows only linearly, so once the plate has moved far enough
the spring can no longer hold it. Stability fails where the two force
slopes match, $\dd F_e/\dd g = -\dd F_s/\dd g$, which after
differentiating and using the equilibrium condition gives the critical
gap at exactly two-thirds of the original,
\[
g_{\text{pi}} = \tfrac{2}{3}g_0.
\]
Substituting back yields the \engterm{pull-in voltage}
\[
V_{\text{pi}} = \sqrt{\frac{8\kappa g_0^{3}}{27\varepsilon_0 A}},
\]
beyond which no equilibrium exists and the plate collapses onto the fixed
electrode. For a micromirror plate of area $A = 100\times 100\,\si{\micro\meter\squared}
= 10^{-8}\,\si{\meter\squared}$, suspension stiffness
$\kappa = 0.5\,\si{\newton\per\meter}$, and rest gap $g_0 = 2\,\si{\micro
\meter}$,
\[
V_{\text{pi}} = \sqrt{\frac{8(0.5)(2\times 10^{-6})^{3}}{27(8.85\times
10^{-12})(10^{-8})}} = 3.7\,\si{\volt},
\]
a voltage a logic supply can reach, which is why electrostatic
micromirrors and switches actuate at single-digit volts. The designer who
wants analogue positioning over the full gap rather than a snap must
operate below one-third of the travel, where the equilibrium is stable, or
add closed-loop control or a series capacitor to extend the stable range
past the natural two-thirds limit.

\subsection{Charge relaxation: how fast a conductor neutralises}

Continuity and Ohm's law together set the time a conductor takes to expel
a charge placed in its interior, the number behind static-dissipative
flooring and the grounding time of a fuelling line. Inside a homogeneous
conductor of conductivity $\sigma$ and permittivity $\varepsilon$,
combine continuity $\partial\rho/\partial t + \divg\vect{J} = 0$, the
field-form Ohm's law $\vect{J} = \sigma\vect{E}$, and Gauss's law
$\divg\vect{E} = \rho/\varepsilon$. Substituting the second and third
into the first,
\[
\pdv{\rho}{t} + \sigma\,\divg\vect{E} = \pdv{\rho}{t} +
\frac{\sigma}{\varepsilon}\rho = 0,
\]
a first-order equation whose solution decays exponentially,
\[
\rho(t) = \rho_0\,e^{-t/\tau}, \qquad \tau = \frac{\varepsilon}{\sigma},
\]
with the \engterm{charge-relaxation time} $\tau = \varepsilon/\sigma$.
Any net charge in the bulk migrates to the surface with this time
constant, leaving the interior neutral, which is the dynamic statement
behind the static result that a conductor carries no interior field. For
copper, $\tau = \varepsilon_0/\sigma = (8.85\times 10^{-12})/(5.96\times
10^{7}) = 1.5\times 10^{-19}\,\si{\second}$, faster than any process an
engineer drives, so a metal is neutral the instant it is charged. For a
good insulator the same ratio is enormous: window glass with $\sigma\sim
10^{-12}\,\si{\siemens\per\meter}$ and $\varepsilon_r\approx 5$ has
$\tau = (5)(8.85\times 10^{-12})/10^{-12} \approx 44\,\si{\second}$, so
charge sits on glass for the better part of a minute. The control of
static discharge in electronics handling and in fuel transfer rests on
choosing materials with $\tau$ short enough to bleed charge away before it
builds to a sparking voltage but long enough not to short the working
circuit; static-dissipative materials are engineered to a relaxation time
in the range of milliseconds to seconds, between the metal and the
insulator. The straight line of \cref{fig:vol04ch08:charge-relaxation}
places all three regimes on one plot: copper at the far right neutralising
in under an attosecond, glass at the far left holding charge for tens of
seconds, and the engineered window in between where the flooring and wrist
straps of an electronics bench and the bonding of a fuelling line are
specified. The design target is a relaxation time short against the process
that builds charge, the walk across a floor or the flow of fuel through a
pipe, and long against the working currents of the circuit the material
touches, a window of a few orders of magnitude that the conductivity alone
selects because the permittivity of solids varies over barely one decade
while the conductivity varies over twenty-four.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-charge-relaxation.tex}

\subsection{Force on a dipole in a gradient: magnetic separation}

A uniform field exerts no net force on a magnetic dipole, only a torque
that aligns it. The force that pulls iron filings toward a magnet, sorts
ore on a conveyor, and localises the signal in a magnetic-resonance scan
needs a field that varies in space, and its size is set by the gradient
rather than by the field itself. A small dipole of moment $\vect{m}$ sitting
in a field $\vect{B}(\vect{r})$ has potential energy $U = -\vect{m}\cdot
\vect{B}$, and the force is the negative gradient of that energy,
\[
\vect{F} = \grad(\vect{m}\cdot\vect{B}).
\]
For a moment held parallel to a field that points along one axis and
changes along it, the force reduces to $F = m\,\dd B/\dd x$: no gradient,
no force. A paramagnetic particle acquires an induced moment $m = \chi V
B/\mu_0$ proportional to the field it sits in, with volume $V$ and volume
susceptibility $\chi$, so its force scales as the product $B\,\dd B/\dd x =
\tfrac12\,\dd(B^2)/\dd x$, the quantity a separator maximises by shaping
its pole faces to a sharp edge. Take a mineral grain of volume $V = 1\,\si{
\milli\meter\cubed} = 10^{-9}\,\si{\meter\cubed}$ and susceptibility
$\chi = 10^{-3}$ in a field $B = 0.6\,\si{\tesla}$ falling over a decay
length $L = 8\,\si{\milli\meter}$, so the gradient is $\dd B/\dd x = B/L =
75\,\si{\tesla\per\meter}$ at the pole edge of
\cref{fig:vol04ch08:dipole-gradient-force}. The moment is $m = \chi V B/
\mu_0 = (10^{-3})(10^{-9})(0.6)/(4\pi\times 10^{-7}) = 4.8\times 10^{-7}\,
\si{\ampere\meter\squared}$, and the force is
\[
F = m\dv{B}{x} = (4.8\times 10^{-7})(75) = 3.6\times 10^{-5}\,\si{\newton},
\]
about $1.2$ times the grain's own weight of $\rho V g \approx 2.9\times
10^{-5}\,\si{\newton}$ for a rock of density $3000\,\si{\kilogram\per\meter
\cubed}$, which is why a high-gradient magnetic separator lifts weakly
paramagnetic mineral against gravity while a diamagnetic gangue with $\chi <
0$ feels a force the other way and stays on the belt. The gradient coils of
a magnetic-resonance scanner apply the same principle to spins rather than
grains, imposing a controlled $\dd B/\dd x$ so the resonant frequency
encodes position, and the force result here is the reason those coils must
be braced against large mechanical loads while they switch.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-dipole-gradient-force.tex}

\subsection{Force on a dielectric slab drawn into a capacitor}

A capacitor pulls a dielectric slab into the gap between its plates, the
force behind a liquid-level gauge that reads the height of oil rising
between two electrodes and behind the variable capacitor of an old radio
tuner. The force is found by the energy method, taking care over which
quantity is held fixed. A parallel-plate capacitor of plate width $w$,
depth into the page $\ell$, and gap $d$ has a slab of permittivity
$\varepsilon = \varepsilon_r\varepsilon_0$ inserted a distance $x$ into the
gap, so the capacitance is the two regions in parallel,
\[
C(x) = \frac{\varepsilon_0 w}{d}\bigl[(\ell - x) + \varepsilon_r x\bigr] =
\frac{\varepsilon_0 w}{d}\bigl[\ell + (\varepsilon_r - 1)x\bigr].
\]
At fixed voltage $V$ the battery does work as the slab moves, so the force
on the slab is the positive gradient of the energy at constant voltage,
\[
F = +\frac{1}{2}V^{2}\dv{C}{x} = \frac{\varepsilon_0 w V^{2}(\varepsilon_r
- 1)}{2 d},
\]
independent of how far the slab has entered, and always positive for
$\varepsilon_r > 1$, so the slab is pulled in. For plates of width $w =
5\,\si{\centi\meter}$ and gap $d = 1\,\si{\milli\meter}$ at $V = 1\,\si{
\kilo\volt}$ with a glass slab of $\varepsilon_r = 5$,
\[
F = \frac{(8.85\times 10^{-12})(0.05)(10^{3})^{2}(5 - 1)}{2(10^{-3})} =
8.9\times 10^{-4}\,\si{\newton},
\]
a fraction of a millinewton, enough to raise a light dielectric liquid a
measurable height and the physical reason a dielectric is drawn to where
the field is strongest. The sign trap is the one the factor-of-one-half
section warned of: at fixed charge the force is $F = -\tfrac12(Q^2/C^2)
\,\dd C/\dd x$, still pulling the slab in but now because the field energy
falls as the higher-permittivity material lowers the field, whereas at
fixed voltage the battery supplies the extra energy and more. Both routes
give the same force at the instant the slab position and the stored charge
coincide; only the accounting differs.

\subsection{The toroid: an inductor that keeps its flux to itself}

A solenoid leaks flux out its ends; a toroid closes the flux path on itself
and confines the field almost entirely to its interior, which is why a
toroidal inductor couples far less noise into its neighbours than a
solenoid of the same inductance. Amp\`ere's law on a circle of radius $r$
threaded through the hole of a toroid of $N$ turns carrying current $I$
encloses all $N$ turns once, so
\[
B(r)\,2\pi r = \mu_0 N I, \qquad B(r) = \frac{\mu_0 N I}{2\pi r},
\]
a field that lives inside the winding and varies across the bore as $1/r$,
stronger at the inner wall than the outer. Outside the toroid any Amp\`erian
loop encloses zero net current, because the turns carry current in and back
out through the same loop, so the exterior field is nominally zero and the
device is self-shielding. The inductance follows from the flux linkage. For
a toroid of rectangular cross-section, inner radius $a$, outer radius $b$,
and height $h$, the flux through one turn is the integral of $B$ across the
face,
\[
\Phi = \int_a^b \frac{\mu_0 N I}{2\pi r}\,h\,\dd r = \frac{\mu_0 N I h}{
2\pi}\ln\frac{b}{a},
\]
and the inductance is $L = N\Phi/I$,
\[
L = \frac{\mu_0 N^{2} h}{2\pi}\ln\frac{b}{a}.
\]
A ferrite toroid with $\mu_r = 2000$, $N = 100$ turns, height $h = 10\,\si{
\milli\meter}$, and radii $a = 10\,\si{\milli\meter}$, $b = 15\,\si{\milli
\meter}$ gives
\[
L = \frac{(2000)(4\pi\times 10^{-7})(100)^{2}(0.010)}{2\pi}\ln\frac{15}{10}
= 1.6\times 10^{-2}\,\si{\henry} = 16\,\si{\milli\henry},
\]
a large inductance in a small closed core, with the permeability of the
ferrite multiplying the vacuum result. The self-shielding is the reason
switching-supply designers reach for a toroid where board space is crowded
and a solenoid's stray flux would trip a nearby sense line, the same
leakage flux the no-monopole section named as the designer's adversary,
here routed almost entirely back into the core.

\subsection{The relay coil, the freewheeling diode, and the L over R clock}

An inductor resists a change in its own current, and interrupting that
current fast forces a voltage spike that destroys the switch unless a path
is provided for the current to decay. A relay coil of inductance $L =
0.5\,\si{\henry}$ and winding resistance $R = 100\,\si{\ohm}$ driven from a
$24\,\si{\volt}$ supply reaches a steady current $I_0 = V/R = 0.24\,\si{
\ampere}$ with time constant $\tau = L/R = 0.005\,\si{\second}$, so it pulls
in within a few milliseconds. When the drive transistor switches off, the
coil current cannot stop instantly; if the only path is the open switch,
the current drives $v = L\,\dd i/\dd t$ to whatever voltage the small stray
capacitance and the breakdown of the switch allow, easily hundreds of volts
across a device rated for tens. A freewheeling diode placed across the coil,
reverse-biased while the coil is energised, becomes forward-biased the
instant the switch opens and carries the coil current in a decaying loop,
clamping the spike to one diode drop above the supply. The current then
decays as
\[
i(t) = I_0\,e^{-t/\tau}, \qquad \tau = \frac{L}{R},
\]
falling to a few percent of $I_0$ in about $3\tau = 15\,\si{\milli\second}$,
which sets how fast the relay can drop out and therefore how fast it can be
cycled. The energy the diode dissipates on each turn-off is the full stored
$\tfrac12 L I_0^{2} = \tfrac12(0.5)(0.24)^{2} = 1.4\times 10^{-2}\,\si{
\joule}$, small per event but a heat load in a relay switched thousands of
times a minute. The design tension is that the same diode that saves the
switch slows the drop-out, so a fast relay uses a resistor or a Zener in
series with the diode to burn the energy faster at a higher, but still
bounded, clamp voltage, trading switch stress against release speed.

\subsection{Magnetic shielding: the mu-metal box around a sensitive probe}

A sensitive magnetometer or a photomultiplier must be kept from the Earth's
field and from stray mains-frequency flux, and the tool is a shell of
high-permeability alloy that diverts the flux around the cavity rather than
through it. The shell works by the reluctance argument of the
magnetic-circuit box: its wall offers the flux a low-reluctance path, so
lines that would have crossed the cavity are drawn into the wall and
returned outside. For a spherical shell of radius $a$ and wall thickness
$t \ll a$ made of material with relative permeability $\mu_r \gg 1$, the
ratio of the external field to the field left inside is the shielding factor
\[
S = \frac{B_{\text{out}}}{B_{\text{in}}} \approx 1 + \frac{2}{3}\mu_r
\frac{t}{a},
\]
the curve of \cref{fig:vol04ch08:magnetic-shielding}. A mu-metal shell of
$\mu_r = 20000$, radius $a = 10\,\si{\centi\meter}$, and wall $t = 3\,\si{
\milli\meter}$ (so $t/a = 0.03$) gives
\[
S \approx 1 + \tfrac{2}{3}(20000)(0.03) = 401,
\]
cutting a $50\,\si{\micro\tesla}$ ambient field to about $0.12\,\si{\micro
\tesla}$ inside. Two design consequences follow from the form of $S$. The
factor grows with the product $\mu_r t$, so a thin wall of a very
high-permeability alloy beats a thick wall of ordinary steel, and mu-metal
earns its cost. The factor also collapses if the wall saturates, because
$\mu_r$ falls toward one once $B$ in the wall approaches the material's
saturation near $0.8\,\si{\tesla}$, so a single thin shell cannot shield a
strong field. The cure is nested shells: an outer shell of a material that
saturates at high field but has modest permeability knocks the field down
to where an inner high-permeability shell can finish the job without
saturating, and the shielding factors of the shells multiply. A three-shell
magnetically shielded room reaches factors in the tens of thousands by
exactly this staging, each shell working the field it receives from the one
outside it.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-magnetic-shielding.tex}

\subsection{Radiated power from an accelerating charge}

A charge moving at constant velocity carries its fields with it and
radiates nothing; a charge that accelerates sheds energy as electromagnetic
radiation, and the rate is the number behind antenna efficiency, the
radiative loss that limited early particle accelerators, and the thermal
glow of any hot filament. The Larmor result gives the total power a
non-relativistic point charge $q$ radiates when accelerated at magnitude
$a$,
\[
P = \frac{q^{2} a^{2}}{6\pi\varepsilon_0 c^{3}},
\]
quadratic in the acceleration, so a charge whipped hard radiates far more
than one nudged. Consider an electron oscillating in an antenna driven at
angular frequency $\omega$ with amplitude $x_0$, so its peak acceleration is
$a = \omega^{2} x_0$. At a broadcast frequency $f = 100\,\si{\mega\hertz}$,
$\omega = 6.3\times 10^{8}\,\si{\per\second}$, and a modest amplitude
$x_0 = 1\,\si{\milli\meter}$, the peak acceleration is $a = (6.3\times
10^{8})^{2}(10^{-3}) = 4.0\times 10^{14}\,\si{\meter\per\second\squared}$,
and the single electron radiates a peak
\[
P = \frac{(1.6\times 10^{-19})^{2}(4.0\times 10^{14})^{2}}{6\pi(8.85\times
10^{-12})(3\times 10^{8})^{3}} = 9.1\times 10^{-25}\,\si{\watt},
\]
vanishing per electron but multiplied by the $10^{18}$ or so carriers
oscillating in step in the antenna current, which is how a coherent array
of accelerating charges turns microwatts of per-charge radiation into the
kilowatts a transmitter launches. The same formula read as a loss governs
the circular accelerator: a charge on a curved path of radius $r$ at speed
$v$ has centripetal acceleration $a = v^{2}/r$, so a beam bent tightly at
high speed radiates away its energy as synchrotron light, the reason a
high-energy electron ring must be built large to keep $a$ and therefore the
radiated loss down, and the reason the same ring is prized as a bright
light source when the radiation is the product rather than the waste. The
$a^{2}$ scaling is the single fact that sets both the efficiency of the
antenna and the cost of the accelerator.

\subsection{The Faraday disk: a generator whose geometry never moves}

Faraday's law in its moving-conductor form gives a steady voltage from a
rotation that changes no circuit shape, the puzzle that named the homopolar
generator. A conducting disk of radius $a$ spins at angular rate $\omega$
about its axis in a uniform axial field $B$, as in
\cref{fig:vol04ch08:faraday-disk}. A free charge at radius $r$ rides with the
disk at speed $\omega r$ and feels a radial magnetic force per unit charge
$\omega r B$, which drives positive charge outward until an electrostatic
field balances it. The steady EMF between the axle and the rim is the
integral of that force per unit charge along the radius,
\[
\mathcal{E} = \int_0^a \omega r B\,\dd r = \tfrac12 B\omega a^{2}.
\]
A disk of radius $a = 0.15\,\si{\meter}$ spinning at $3000$ revolutions per
minute, $\omega = 314\,\si{\radian\per\second}$, in a field $B = 0.5\,\si{
\tesla}$ generates
\[
\mathcal{E} = \tfrac12(0.5)(314)(0.15)^{2} = 1.8\,\si{\volt},
\]
a low voltage at a potentially enormous current, since the single-turn disk
has almost no resistance. The output does not reverse and needs no
commutator, because every radial element sweeps flux in the same sense
throughout the turn, which is what makes the homopolar machine the natural
source for the megaampere pulses the electromagnetic launcher of the fourth
vignette demands. The puzzle that troubled its early users, that the flux
through any fixed circuit built from the disk and its brushes never changes
while the disk still generates, is resolved by the moving-conductor form:
the EMF is the line integral of $\vect{v}\times\vect{B}$ along the moving
material of the disk, not $-\dd\Phi/\dd t$ through a stationary loop, and the
two forms agree only when the circuit's material is at rest.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-faraday-disk.tex}

\subsection{The Helmholtz pair: a flat field on a bench}

A single coil produces a field that peaks in its plane and falls off sharply,
useless for a sample that needs a uniform field over a finite volume. Two
identical coaxial coils of radius $a$ carrying equal currents in the same
sense, spaced a distance $s$ apart, superpose their on-axis fields, and a
particular spacing flattens the sum. Each coil of $N$ turns carrying $I$
produces on its axis at axial distance $z$ from the coil plane the field
$B(z) = \mu_0 N I a^{2}/[2(a^{2}+z^{2})^{3/2}]$, the Biot-Savart result for a
loop. Place the coils at $z = \pm s/2$ and expand the sum about the midpoint.
By symmetry the first and third derivatives vanish there; the second
derivative vanishes as well when the spacing equals the radius,
\[
s = a,
\]
the \engterm{Helmholtz condition}, leaving the field uniform to better than
one percent over a central region roughly a third of the radius wide, the
blue plateau of \cref{fig:vol04ch08:helmholtz}. The centre field of the pair
is
\[
B_{\text{centre}} = \frac{\mu_0 N I a^{2}}{[a^{2}+(a/2)^{2}]^{3/2}}\cdot 1
= \left(\frac{4}{5}\right)^{3/2}\frac{\mu_0 N I}{a} = 0.716\,\frac{\mu_0 N I}{
a}.
\]
A pair of $N = 130$ turn coils of radius $a = 0.20\,\si{\meter}$ carrying
$I = 1.0\,\si{\ampere}$ produces at the centre
\[
B_{\text{centre}} = 0.716\,\frac{(4\pi\times 10^{-7})(130)(1.0)}{0.20} =
5.8\times 10^{-4}\,\si{\tesla} = 0.58\,\si{\milli\tesla},
\]
about ten times the Earth's field, enough to cancel the geomagnetic
background over a sample and reverse it under control. Spacing the coils too
far apart splits the plateau into two humps with a dip at the centre, the
amber curve of the figure, the sign that the uniformity condition has been
overshot; spacing them too close leaves a single rounded peak. The Helmholtz
geometry is the standard way to make a known, uniform field on a bench, and
the calibration of the chapter's project magnetometer is done inside one.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-helmholtz.tex}

\subsection{Poynting flux into a resistor: energy enters through the sides}

The principle that energy travels in the fields, stated for the coaxial cable,
is sharpest on a single current-carrying resistor, where the Poynting vector
points radially inward through the wire's surface and delivers exactly the
dissipated power. Take a cylindrical wire of radius $a$ and length $L$
carrying steady current $I$ against resistance $R$, so a voltage $V = IR$
falls along it. At the surface the electric field is axial, set by the
voltage gradient, $E = V/L$, parallel to the current; the magnetic field
circles the wire, $B = \mu_0 I/(2\pi a)$ from Amp\`ere's law at the surface.
The Poynting vector $\vect{S} = \vect{E}\times\vect{B}/\mu_0$ is therefore
radial and points \emph{inward}, with magnitude
\[
S = \frac{E B}{\mu_0} = \frac{V}{L}\cdot\frac{I}{2\pi a},
\]
and the total power flowing in through the cylindrical surface of area
$2\pi a L$ is
\[
P_{\text{in}} = S\,(2\pi a L) = \frac{V}{L}\cdot\frac{I}{2\pi a}\cdot 2\pi a L
= VI = I^{2}R,
\]
exactly the dissipated power. A $10\,\si{\ohm}$ resistor carrying $2\,\si{
\ampere}$ dissipates $I^{2}R = 40\,\si{\watt}$, and that $40\,\si{\watt}$
enters through the resistor's cylindrical surface as radial field energy, not
along the wire's axis with the current. The energy the load consumes comes
from the space around the conductor, guided in through the sides by the
crossed fields, and the wire is the sink that converts it to heat rather than
the pipe that carries it. This is the same accounting that gave the coaxial
cable its axial power flux, run on the simplest possible circuit, and it is
the reason the field, not the current, is the honest carrier of
electromagnetic energy.

\subsection{The Rogowski coil: measuring a current without breaking it}

Amp\`ere's law measures a current from the field it makes, and a coil wound
as a flexible loop around a conductor turns that measurement into a
clip-on instrument that reads even large currents without cutting the wire. A
\engterm{Rogowski coil} is a helical winding of $n$ turns per unit length and
small cross-section $A$ wound uniformly along a closed loop that encircles the
conductor. The changing field of the enclosed current $I(t)$ induces in the
winding a voltage equal to the rate of change of the linked flux. Summing the
flux linked by each turn around the loop and applying Amp\`ere's law
$\oint\vect{B}\cdot\dd\vect{\ell} = \mu_0 I$ to the coil's centreline gives an
output voltage
\[
v(t) = -\mu_0 n A\,\dv{I}{t},
\]
independent of where the conductor sits inside the loop and of the loop's
shape, because the uniform winding density makes the sum depend only on the
total enclosed current. The output is proportional to $\dd I/\dd t$, so the
instrument integrates it, in an analog $RC$ integrator or in software, to
recover $I(t)$. For a coil of $n = 2000\,\si{\per\meter}$ turns and
cross-section $A = 1\,\si{\centi\meter\squared} = 10^{-4}\,\si{\meter
\squared}$ measuring a current rising at $\dd I/\dd t = 10^{6}\,\si{\ampere
\per\second}$, a typical power-electronics switching rate, the raw output is
\[
v = (4\pi\times 10^{-7})(2000)(10^{-4})(10^{6}) = 0.25\,\si{\volt},
\]
a clean signal at the coil terminals before integration. The Rogowski coil
has no iron core to saturate, so it reads currents from milliamperes to
hundreds of kiloamperes with the same coil and stays linear where a
current transformer's core would have long since saturated, which is why it
is the tool for lightning-current and fault-current capture as well as for
routine power measurement. Its blind spot is direct current: with $\dd I/
\dd t = 0$ the output is zero, so it reads only changing currents, the price
of coupling through Faraday's law rather than through a Hall element.

\subsection{Radiation resistance of a short dipole antenna}

The Larmor result gives the power a single accelerating charge radiates; the
same physics, summed over the charges oscillating in an antenna, appears at
the terminals as a resistance that carries no heat but represents power
launched into space. A short dipole of length $\ell$ much smaller than the
wavelength $\lambda$, driven with a current of amplitude $I_0$ that tapers to
zero at its ends, radiates a time-averaged power that works out to
\[
P_{\text{rad}} = \frac{\pi}{3}\eta_0\left(\frac{\ell}{\lambda}\right)^{2}
\frac{I_0^{2}}{4} = \tfrac12 I_0^{2} R_{\text{rad}}, \qquad
R_{\text{rad}} = 20\pi^{2}\left(\frac{\ell}{\lambda}\right)^{2},
\]
where $\eta_0 = 377\,\si{\ohm}$ is the impedance of free space and the
\engterm{radiation resistance} $R_{\text{rad}}$ is the equivalent resistance
that would dissipate the same power the antenna radiates. The quadratic
$(\ell/\lambda)^{2}$ dependence is the Larmor $a^{2}$ scaling in disguise,
since a shorter antenna at a fixed frequency holds its charges to smaller
excursions and a smaller acceleration. A $1\,\si{\meter}$ whip driven at
$1\,\si{\mega\hertz}$, where $\lambda = 300\,\si{\meter}$, has
\[
R_{\text{rad}} = 20\pi^{2}\left(\frac{1}{300}\right)^{2} = 2.2\times 10^{-3}
\,\si{\ohm},
\]
a couple of milliohms, far below the ohm or so of ohmic loss in the whip and
its ground connection, so the antenna radiates only a small fraction of the
power fed to it and is a poor radiator. Raise the frequency to $100\,\si{
\mega\hertz}$, where $\lambda = 3\,\si{\meter}$ and the same whip is a third
of a wavelength, and $R_{\text{rad}}$ climbs by the square of the frequency
ratio to about $22\,\si{\ohm}$, comparable to the feed impedance, and the
antenna radiates efficiently. The lesson matches the failure section's
principle that every cable is an antenna: a conductor short against the
wavelength radiates almost nothing, which protects a low-frequency circuit and
dooms a low-frequency antenna, while the same conductor at a frequency where
it spans a fair fraction of a wavelength radiates in earnest, whether the
designer wanted it to or not.

\subsection{Shielding effectiveness: reflection plus absorption}

The skin-depth box set the thickness a shield needs; the full accounting adds
a second mechanism and gives the shield a number in decibels. A conductive
enclosure blocks an external field two ways, plotted against frequency in
\cref{fig:vol04ch08:shielding-effectiveness}. The first is reflection: the
wave in air, with impedance $\eta_0 = 377\,\si{\ohm}$, meets the conductor's
tiny surface impedance $\eta_s = \sqrt{\mu\omega/(2\sigma)}\ll\eta_0$, and the
mismatch turns most of the wave back, exactly the interface-condition
reflection of the dielectric example carried to a conductor. The reflection
loss for a plane wave is
\[
R \approx 168 - 10\log_{10}\!\left(\frac{f\,\mu_r}{\sigma_r}\right)\,\si{
\decibel},
\]
with $\mu_r$ and $\sigma_r$ relative to copper, falling as frequency rises
because the conductor's surface impedance grows with $\sqrt{f}$ and the
mismatch narrows. The second mechanism is absorption: the fraction of the
wave that does enter is attenuated across the wall as $e^{-t/\delta}$, giving
an absorption loss
\[
A = 8.686\,\frac{t}{\delta} = 8.686\,t\sqrt{\pi f\mu\sigma}\,\si{\decibel},
\]
rising as $\sqrt{f}$ because the skin depth shrinks. For a copper sheet
$t = 0.5\,\si{\milli\meter}$ thick at $1\,\si{\mega\hertz}$, the skin depth is
$65\,\si{\micro\meter}$, so $t/\delta = 7.7$ and the absorption loss is
$A = 8.686\times 7.7 = 67\,\si{\decibel}$, while the reflection loss is
$R = 168 - 10\log_{10}(10^{6}) = 108\,\si{\decibel}$, for a total shielding
effectiveness of $175\,\si{\decibel}$, a factor of $10^{8.75}$ in field
amplitude. The two terms trade roles across the band: reflection dominates at
low frequency where the sheet is thin in skin depths, absorption dominates at
high frequency, and their sum dips through a shallow minimum where the two
curves cross, the frequency a designer checks first because it is where the
enclosure is weakest. A gap or a seam undoes all of it, since a slot longer
than a fraction of a wavelength leaks like an antenna regardless of the wall's
own effectiveness, which is why a shielded enclosure lives or dies by its
seams, gaskets, and the size of the holes cut for cables.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-shielding-effectiveness.tex}

\begin{mastery}
The scattered force results of this section, the pull on a capacitor plate,
the pinch on a plasma, the repulsion of two busbars, are all one accounting
once the force is written as a flux of momentum through a surface. The
\engterm{Maxwell stress tensor} collects the field's mechanical action into
a single object
\[
T_{ij} = \varepsilon_0\!\left(E_i E_j - \tfrac12\delta_{ij}|\vect{E}|^{2}
\right) + \frac{1}{\mu_0}\!\left(B_i B_j - \tfrac12\delta_{ij}|\vect{B}|^{2}
\right),
\]
whose meaning is that the force on all the charge and current inside a
closed surface equals the integral of $T_{ij}$ over that surface, plus a
term for the momentum stored in the field,
\[
F_i = \oint_{\partial V} T_{ij}\,\dd A_j - \frac{\dd}{\dd t}\int_V
\varepsilon_0(\vect{E}\times\vect{B})_i\,\dd V.
\]
The engineer never has to know the charges inside; the force follows from
the fields on a surface drawn anywhere around them. Read the diagonal terms
and the earlier pressures fall out at once: a field $\vect{E}$ normal to a
surface presses with $\tfrac12\varepsilon_0 E^{2}$, the electrostatic
pressure that pulled the capacitor plate, while a field $\vect{B}$ tangent
to a surface pushes with $B^{2}/(2\mu_0)$, the magnetic pressure that
pinched the plasma and burst the high-field winding. The off-diagonal terms
carry the shear that twists a motor's rotor and drags an eddy-current
brake. The stress tensor is the field theory's statement that momentum, like
energy, lives in the field and flows through space: the last term above is
the field momentum whose density is the Poynting vector over $c^{2}$, the
same momentum the radiation-pressure exercise delivers to an absorbing
surface. Every force in the chapter is one evaluation of this one surface
integral, which is why the field, not the wire and not the charge, is the
right accounting unit for electromagnetic force.
\end{mastery}

\begin{mastery}
The induced electric field of the changing-flux example exposes a
distinction the static reader can miss: the electric field splits into two
pieces that obey different rules. Write $\vect{E} = -\grad\phi -
\partial_t\vect{A}$ from the potential decomposition. The first piece,
$\vect{E}_{\text{C}} = -\grad\phi$, is the \engterm{conservative} or
Coulomb field; its curl is zero, its line integral around any closed loop
is zero, and it is the gradient of the scalar potential that voltmeters
measuring electrostatic potential read. The second piece,
$\vect{E}_{\text{i}} = -\partial_t\vect{A}$, is the \engterm{induced}
field; its curl is $-\partial_t\vect{B}$, and its line integral around a
loop is the EMF of Faraday's law, nonzero whenever flux changes. Only the
sum is the physical field that pushes a charge. The split explains why
``voltage'' is ambiguous in a circuit with changing flux: the integral
$\int\vect{E}\cdot\dd\vect{\ell}$ between two points depends on the path
taken, because the induced part is not a gradient, so two voltmeters
across the same two nodes can read different values if their leads enclose
different flux. The lumped-circuit fiction that a node has a single
well-defined potential is exactly the quasi-static approximation that the
induced field is negligible inside the wires and confined to the
inductors, where it is bookkept as $L\,\dd i/\dd t$. When that
approximation fails, at high frequency or near a strong changing flux, the
engineer must return to the field and abandon the single-valued node
voltage.
\end{mastery}

\section{An applied vignette: an inductive wireless-charging pad}\pathtag{standard}

A phone-charging pad transfers power across an air gap with no contact,
and it is Faraday's law and the mutual-inductance result of the worked
examples turned into a consumer product. A transmitter coil in the pad
carries an alternating current; the changing flux it produces links a
receiver coil in the phone; the induced EMF drives current into the
phone's battery charger. The whole device is two coupled inductors, and
its behaviour is read off the coupling coefficient and the resonance the
designer tunes.

The transmitter coil of $N_1$ turns carries a current at frequency $f$,
chosen in the band near $100$ to $200\,\si{\kilo\hertz}$ for the common
open standard, high enough that a modest mutual inductance couples useful
power and low enough that switching losses stay manageable. The flux it
links into a coaxial receiver coil of $N_2$ turns at axial gap $z$ sets
the mutual inductance $M$, and the coupling coefficient $k = M/\sqrt{L_1
L_2}$ measures how completely the two share flux. For thin coaxial coils
of equal radius $a$ the coupling falls steeply with gap,
\[
k \approx \frac{1}{\left[1 + (z/a)^{2}\right]^{3/2}},
\]
the curve of \cref{fig:vol04ch08:wireless-charging}: at a gap equal to the
coil radius the coupling has already dropped to about a third, and the
transferred power with it. A practical pad holds the gap below half the
coil radius, where $k$ stays above about $0.7$, which is why the phone
must lie flat and aligned on the pad and why misplacing it stops the
charge. The receiver coil is made as large as the phone body allows, both
to raise $a$ so the gap-to-radius ratio stays small and to intercept more
flux.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-wireless-charging.tex}

Loose coupling alone would waste most of the power, so the pad does not
run the coils as bare inductors. Each coil is tuned with a series or
parallel capacitor to resonate at the operating frequency, $f = 1/(2\pi
\sqrt{LC})$, so the reactive voltage the loosely coupled inductor would
demand is supplied by the resonant capacitor rather than by the source.
At resonance the coil current, and the flux it produces, is larger than
the source could drive into the bare inductor by the quality factor $Q =
\omega L/R$ of the resonant coil, so a pad with $k$ as low as $0.3$ can
still deliver useful power if both coils have $Q$ in the tens. The
power transferred scales as $k^{2}Q_1 Q_2$ in the figure of merit that
governs every loosely coupled resonant link, so the designer who cannot
raise $k$ raises $Q$ instead, trading bandwidth for transfer at a fixed
geometry. The combination of moderate coupling and high $Q$ is what makes
contactless charging work across the few-millimetre gap of a phone case.

The curve of \cref{fig:vol04ch08:resonant-transfer} makes the trade
concrete. The peak efficiency of an optimally loaded pair depends on the
coupling and the quality factors only through the single figure of merit
$U^{2} = k^{2}Q_1 Q_2$, and it climbs toward one as that product grows. A
pad with a third of the flux coupled and quality factors of forty reaches a
figure of merit near $144$ and a transfer efficiency around seventy
percent, while a poorly coupled link of quality factor ten and coupling a
tenth reaches a figure of merit of one and delivers under a fifth of the
available power. The lesson the curve carries into every loosely coupled
system, from a charging pad to a medical implant's power link, is that the
coupling and the quality factors are interchangeable within the product:
what geometry loses to a wide gap, resonance can recover with a sharp tank,
up to the bandwidth the sharpness costs.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-resonant-transfer.tex}

The efficiency that reaches the battery is the product of the coil-to-coil
transfer and the losses in the switching electronics on both sides. A
typical pad delivers $5$ to $15\,\si{\watt}$ to the phone at an
end-to-end efficiency near seventy percent, well below a wired charger's
ninety-plus, with the difference appearing as heat in the coils and the
power electronics, the same resistive $\vect{J}\cdot\vect{E}$ dissipation
that heated the transformer windings. The lost power warms both the pad
and the phone, which is why a wirelessly charging phone runs warmer than a
wired one and why the pad limits its power when it detects a rising
temperature. A foreign metal object left on the pad is the safety hazard
the design must catch: the transmitter's changing flux drives eddy
currents in any conductor lying on it, heating a coin or a key as an
eddy-current brake heats its sheet, so the pad monitors the reflected
load and shuts down when the power drawn does not match a coupled receiver,
the foreign-object detection every compliant pad carries.

\section{A second vignette: the induction cooktop}\pathtag{standard}

The wireless-charging pad transferred power across a gap to a matched
receiver; an induction cooktop uses the same coupled-coil physics with the
opposite intent, dumping power as heat into the load rather than delivering
it intact, and the load is the cooking pan itself. A flat spiral coil under
the ceramic top carries an alternating current near $25\,\si{\kilo\hertz}$;
the changing flux it produces threads the ferromagnetic base of the pan;
the induced eddy currents dissipate against the pan's own resistance, and
the pan heats from within while the glass top and the coil stay
comparatively cool. The whole appliance is Faraday's law, the skin effect,
and the eddy-current dissipation of the chapter turned into a stove.

The power deposited in the pan is the eddy-current loss of the moving-sheet
example, adapted to a fixed sheet in an alternating field. The induced field
in the pan base drives a current density $J \sim \sigma\omega B\delta$
confined to a surface layer one skin depth $\delta = \sqrt{2/(\mu\sigma
\omega)}$ deep, and the volumetric dissipation $\vect{J}\cdot\vect{E} =
J^{2}/\sigma$ integrated over that layer gives a surface power density that
rises with the drive frequency and with the pan's surface resistance. The
material choice is what makes the appliance work.
\cref{fig:vol04ch08:induction-cooktop} plots the skin depth against
frequency for a ferromagnetic steel pan and an aluminium one. Steel wins
twice over: its permeability $\mu_r$ of a few hundred shrinks the skin depth
by $\sqrt{\mu_r}$ relative to a non-magnetic metal, crowding the current
into a thin high-resistance shell, and that thin shell dissipates more of
the drive energy as heat than aluminium's thicker, lower-resistance layer.
At the working frequency the steel skin depth is a fraction of a millimetre,
well inside a $2\,\si{\milli\meter}$ pan base, while the aluminium skin
depth is a large fraction of the base, so the aluminium pan lets the field
pass and couples poorly. A bare aluminium or copper pan barely heats on an
induction top, which is why induction cookware carries a ferromagnetic base
layer even when the body is aluminium, and why a kitchen magnet sticking to
the base is the quick test of compatibility.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-induction-cooktop.tex}

The frequency choice threads two constraints. It sits above the audible
band, near $20$ to $30\,\si{\kilo\hertz}$, so the magnetostrictive hum of
the pan and the coil is inaudible or nearly so, and it sits below the range
where switching losses in the drive electronics and radiated emission would
grow costly. The coil is driven by a resonant inverter, the pan-and-coil
combination tuned so the drive sees a near-resistive load at the operating
frequency, the same resonance trick the charging pad used to force useful
current through a loosely coupled inductor. The coupling to the pan is far
tighter than the pad's coupling to a phone, because the pan sits directly on
the ceramic with a gap of a few millimetres and covers the coil, so a large
fraction of the flux links the load and end-to-end efficiencies reach
eighty-five to ninety percent, above the wireless pad and close to a
resistive element, with the advantage that the heat is generated in the pan
rather than in a glowing element that must then conduct it upward.

The safety behaviour falls out of the same coupling. With no pan present the
coil sees only air, the reflected load is small, and the drive detects the
mismatch and idles rather than radiating into an empty top, the same
reflected-load sensing the charging pad used for foreign-object detection.
A small ferromagnetic object, a dropped knife or a ring, couples enough to
heat locally, so the appliance sets a minimum detectable load and refuses
to run for anything smaller than a pan-sized footprint. The stray field
above the top falls off steeply beyond the pan edge because the pan, like
the toroid's return path, largely closes the flux, so the field a bystander
stands in is a small fraction of the field under the pan, within the
exposure limits the appliance is certified against. The induction cooktop
is the chapter's eddy-current, skin-effect, and coupled-coil results
assembled into a device that heats by loss on purpose, the mirror image of
the transformer that fought every watt of the same loss.

\section{A third vignette: the fluxgate magnetometer}\pathtag{standard}

The Hall sensor of the worked examples reads a field through the Lorentz
force on a moving carrier, and it reaches millitesla resolution; a compass
needle reads the Earth's field but cannot be logged or automated. The
fluxgate magnetometer reads a field a thousand times weaker than the Earth's
by turning the nonlinearity of a saturating core into a frequency doubler,
and it is Faraday's law and the $B$--$H$ saturation curve of the transformer
case study turned into the most widely used vector field sensor in
spacecraft, geophysics, and unexploded-ordnance survey.

The core of the sensor is a small strip or ring of high-permeability alloy
wound with two windings, a drive winding and a sense winding. An
alternating drive current pushes the core hard into saturation in both
directions each half-cycle, so the flux in the core follows the flat-topped
$B$--$H$ curve of \cref{fig:vol04ch08:hysteresis-loop} rather than the drive
sinusoid: the core spends part of each half-cycle saturated, where its
permeability has collapsed and it can accept no more flux. With no external
field the two saturation events per cycle are symmetric, and the voltage the
changing flux induces in the sense winding, $\mathcal{E} = -N\,\dd\Phi/\dd
t$, contains only odd harmonics of the drive frequency. An external field
along the core axis biases the saturation: the core reaches saturation
earlier on the half-cycle that the external field aids and later on the one
it opposes, so the two saturation events are no longer symmetric, and the
asymmetry injects a \engterm{second harmonic} into the sense voltage whose
amplitude is proportional to the external field and whose phase gives its
sign. The instrument drives the core at a frequency $f$, tunes the sense
winding to $2f$, and reads the second-harmonic amplitude as the field.

The physics that makes it sensitive is the steepness of the saturation
knee. A small external field shifts the saturation timing by an amount set
by how sharply the permeability collapses at the knee, so a core with an
abrupt knee and a high unsaturated permeability converts a tiny field into a
large timing asymmetry and a large second harmonic. Permalloy and
amorphous cobalt alloys are chosen for exactly this square-loop character.
A practical fluxgate resolves fields in the nanotesla range, four orders of
magnitude below the Earth's $50\,\si{\micro\tesla}$, and holds a stable
zero over temperature because the reading depends on the timing of a
saturation event rather than on an absolute voltage that would drift. Run in
a feedback loop, with a nulling current fed back through the sense winding to
hold the core at zero net field, the instrument becomes linear and its scale
factor is set by the feedback resistor rather than by the core, the same
trick that linearised the pull-in microactuator by closing a loop around an
unstable open-loop device.

The fluxgate's limits trace to the same core. It measures the field
component along its axis only, so a vector instrument uses three orthogonal
cores; it saturates and reads nothing in a field strong enough to hold the
core saturated through the whole cycle, so it is a low-field instrument that
a strong magnet blinds; and its own drive current radiates at the drive
frequency, so a magnetometer array spaces its sensors and staggers their
drives to keep each from reading its neighbour. A spacecraft flies its
fluxgate on a boom for the same reason, holding the sensor several metres
from the currents and motors of the bus whose stray fields, dying as the
dipole $r^{-3}$ of the multipole ladder, would otherwise swamp the faint
interplanetary field the mission was built to measure.

\section{A fourth vignette: the electromagnetic launcher}\pathtag{standard}

The solenoid actuator pulled a plunger a few millimetres; the electromagnetic
launcher applies the same force law over a metre-long barrel to throw a
projectile to kilometres per second, and it is the Lorentz force, the
magnetic pressure, and the $L\,\dd i/\dd t$ energy of the chapter driven to
their violent extreme. Two forms share the physics. A \engterm{railgun}
runs a large current up one rail, across a sliding conductive armature, and
back down the other rail; a \engterm{coilgun} pulls a ferromagnetic or
conducting slug through a series of coils switched in sequence. Both convert
stored electrical energy into kinetic energy through a force that is the
chapter's force-on-a-current in a self-generated field.

In the railgun the current $I$ flows through the armature sitting in the
magnetic field its own rails produce, and the force on the armature is the
same $F = B I \ell$ that drove the rail-and-rod generator of Section 8.8,
now with the field supplied by the rails rather than by an external magnet.
Writing the force in terms of the rails' inductance gradient makes it a
clean energy statement: as the armature slides a distance $\dd x$ the
circuit inductance rises by $L'\,\dd x$ with $L'$ the inductance per unit
length of the rail pair, and the force is
\[
F = \tfrac12 L' I^{2},
\]
the same energy-gradient form that gave the solenoid actuator its pull,
here with the gradient taken along the barrel rather than across an air gap.
A rail pair has $L'$ of order $0.5\,\si{\micro\henry\per\meter}$, so a
current of $I = 10^{6}\,\si{\ampere}$, a megaampere, gives
\[
F = \tfrac12(0.5\times 10^{-6})(10^{6})^{2} = 2.5\times 10^{5}\,\si{\newton},
\]
a quarter meganewton on a projectile of a few kilograms, an acceleration of
tens of thousands of $g$. The current comes from a pulsed-power supply, the
capacitor bank or compensated pulsed alternator of the pulsed-power worked
example, discharged into the rails in a few milliseconds, and the same
critical-damping and voltage-reversal cautions of that example govern the
switching.

The launcher's hard limits are the chapter's failure mechanisms turned up
to full scale. The magnetic pressure $B^{2}/(2\mu_0)$ between the rails
tries to blow them apart, so the rails are massively braced against a force
that, at the megaampere current, reaches the plasma-pinch pressures of the
magnetic-pressure example; the rails must survive it thousands of times.
The sliding contact carries the full current through a moving interface, and
the $\vect{J}\cdot\vect{E}$ dissipation there melts and ablates the rail
surface, the resistive loss of the Poynting balance concentrated into a
running hot spot, which is why rail life is the technology's central
unsolved problem. The armature accelerates so hard that at high speed a
\engterm{velocity skin effect} crowds the current into the trailing edge of
the contact, since the field cannot diffuse into the rail fast enough to
follow the moving armature, the same skin-depth diffusion of the mastery box
now competing against mechanical motion rather than against an oscillation.
The coilgun avoids the sliding contact by coupling to the slug inductively,
trading the rail-erosion problem for the timing problem of switching each
coil at the instant the slug reaches it, a sequencing that a position sensor,
often a Hall probe of the worked examples, must drive. Both forms show the
chapter's lesson at its starkest: the force, the pressure, the loss, and the
energy that a benign transformer balances to a few percent are, in the
launcher, pushed until every one of them becomes the design limit at once.

\section{A fifth vignette: electrodynamic maglev}\pathtag{standard}

The eddy-current brake dissipated a moving conductor's energy into heat; run
the same physics under a vehicle instead of a brake shoe and the drag that
stopped the brake becomes the lift that floats a train. An
\engterm{electrodynamic maglev} carries strong magnets on the vehicle and runs
them over a passive conducting guideway; the moving field induces eddy currents
in the guideway, and those currents produce a field that repels the magnets and
holds the vehicle off the track. The device is the eddy-current, induced-field,
and magnetic-pressure results of the chapter assembled into a suspension that
touches nothing.

The lift and drag both trace to the phase of the induced currents, and the
phase is set by how fast the vehicle moves. A magnet moving at speed $v$ over
a sheet of surface resistance drives eddy currents whose magnitude and phase
follow the sheet's own $L/R$ response to the moving field. At low speed the
induced currents are small and nearly in phase with the driving field, so they
mostly oppose the motion: drag dominates and lift is weak, the rising left
side of \cref{fig:vol04ch08:eddy-maglev}. As the speed climbs past a
characteristic value $v_c$ set by the guideway's resistance and the magnet
geometry, the currents grow and shift toward a quarter-cycle lag, so their
field turns from opposing the motion to repelling the magnet: lift rises
toward an asymptote while drag peaks at $v_c$ and then falls. Above the drag
peak the lift-to-drag ratio grows in direct proportion to speed,
\[
\frac{F_{\text{lift}}}{F_{\text{drag}}} \approx \frac{v}{v_c},
\]
so a vehicle that runs well above $v_c$ floats efficiently while one crawling
below it fights a large drag for little lift. This is why an electrodynamic
maglev rolls on wheels up to a lift-off speed of tens of metres per second and
only rises onto its field once it has passed the drag peak, the opposite of an
electromagnetic (attractive) suspension that levitates from rest but needs the
active feedback the pull-in instability of the worked examples forces.

\input{volumes/04-energy/ch08-electromagnetism-maxwell/figures/fig-eddy-maglev.tex}

The suspension is stable without control, which is its central advantage. A
magnet that sinks toward the guideway induces stronger currents and a stronger
repulsion, so the lift rises to push it back, a passive restoring force the
attractive scheme cannot supply. The cost is the drag, which is pure
dissipation: the power lost to the guideway is $F_{\text{drag}}\,v$, the same
resistive $\vect{J}\cdot\vect{E}$ heating that warmed the eddy-current brake's
sheet, now spread along kilometres of track and paid continuously as the price
of contactless running. A practical system drives $v_c$ down, so the vehicle
spends its cruise far above the drag peak, by making the guideway a good
conductor and by shaping the vehicle field with a Halbach array that
concentrates the flux on the guideway side and cancels it on the cabin side,
raising the coupled field for a given magnet mass. The magnetic pressure
$B^{2}/(2\mu_0)$ of the chapter sets the lift per unit area the same way it set
the actuator's pull: a guideway field of a few tenths of a tesla holds a few
tens of kilopascals, tonnes per square metre, enough to carry a train on a
modest pole area. The maglev is the eddy current read as a feature rather than
a loss, the mirror of the induction cooktop that wanted the loss and the
transformer that fought it.

\begin{mastery}
The failure section's principle that every cable is an antenna carries a
partner the transmit-and-receive symmetry of an antenna, and the partner is a
theorem. Electromagnetic \engterm{reciprocity} states that for two locations
$1$ and $2$ in any medium built from linear, isotropic, non-moving materials,
the voltage a source current at $1$ induces at $2$ equals the voltage the same
source current at $2$ would induce at $1$. Written for two current
distributions $\vect{J}_a$ and $\vect{J}_b$ and the fields $\vect{E}_a$,
$\vect{E}_b$ they produce, the Lorentz reciprocity relation is
\[
\int \vect{J}_a\cdot\vect{E}_b\,\dd V = \int \vect{J}_b\cdot\vect{E}_a\,\dd V,
\]
a consequence of Maxwell's equations that holds whenever the constitutive
parameters are symmetric, which is to say for ordinary dielectrics,
conductors, and non-gyrotropic magnetic materials. The engineering payoff is
that an antenna's radiation pattern, its gain, and its impedance are identical
whether it transmits or receives, so a directional antenna aimed to send a
beam one way is equally deaf to signals from every other way, and a single
measurement of the pattern serves both roles. The theorem also underwrites the
crosstalk symmetry the failure section leans on: if a noisy cable couples a
given fraction of its energy into a victim, the victim driven at the same
frequency couples the same fraction back, so a coupling path is a two-way
street and shielding one end helps the other. Reciprocity fails exactly where
the symmetry of the materials fails, in a magnetised ferrite or a moving
plasma, and that failure is the basis of the isolator and the circulator, the
one-way electromagnetic components a radar or a base station needs to keep a
transmitter's power out of its own receiver. The theorem is thus useful twice
over: it collapses two antenna problems into one wherever it holds, and it
names the special materials an engineer must reach for when a one-way path is
the whole point.
\end{mastery}

\section{Failure: ground loops, induced EMI, lightning}\pathtag{core}

Every electrical system in the world is embedded in an
electromagnetic environment that includes power-line magnetic fields,
radio-frequency signals, and atmospheric electrical activity. Three
failure mechanisms recur across every domain in which fields are
measured or signals are transmitted, and a chapter that does not name
them has not described real electromagnetism.

A \engterm{ground loop} arises when two devices that share a signal
connection also share separate paths to earth. The two earth paths,
typically through separate building grounds, are at slightly different
potentials because of currents flowing through the building's
distribution system. The voltage difference drives a current through
the signal connection between the devices, contaminating any
small-signal measurement that rides on top of it. The mechanism is
Faraday's law applied to the area enclosed by the two ground paths
and the signal cable: time-varying magnetic flux through that area
induces an EMF around the loop. Mains-frequency hum at $50\,\si{\hertz}$
or $60\,\si{\hertz}$ in audio recordings, oscilloscope traces, and
biological signal recordings is the everyday signature.

The cure is a single-point ground: every device's earth connection is
brought to a single physical node, so no loop area exists. When a
single-point ground cannot be achieved (long signal runs, multiple
buildings), the working tools are differential signalling (the signal
is the difference between two conductors, and common-mode interference
on both is rejected), optical isolation (the signal is converted to
light and back, breaking the electrical connection), and transformer
coupling (the signal is transferred through a magnetic core, again
breaking the direct electrical connection).

\engterm{Electromagnetic interference} (EMI) is unintended coupling of
electromagnetic energy from a source into a victim circuit. The source
may be a switching power supply whose high-$dI/dt$ current pulses
radiate; a digital clock line whose square-wave edges have harmonic
content into the gigahertz; a motor whose commutation produces
broadband sparks; a radio transmitter operating nearby. The coupling
mechanism may be capacitive (high-impedance victim sees high-$dV/dt$
source through stray capacitance), inductive (low-impedance victim
sees high-$dI/dt$ source through mutual inductance), or radiative
(the source is far enough away that near-field coupling has fallen
off and the wave propagates as a plane wave; Chapter 10). The
frequency band of the source sets which coupling dominates and which
wavelength matters; the table \repopath{data/em\_spectrum\_bands.csv}
lists the engineering bands from power-line and audio frequencies
through the radio spectrum to visible light, with the wavelength range
of each, so that a conductor's length can be compared against the
wavelength of the interference it carries.

Two principles govern EMI mitigation. First, the source is reduced at
the source: a switching converter's $dI/dt$ is shaped by snubber
networks, slew-rate limits, or spread-spectrum modulation that
spreads the harmonic content over a band rather than concentrating it
at a single frequency. Second, the victim is shielded and the loop
areas of victim circuits are minimised: a twisted pair carrying the
signal and its return places the two conductors close together, so
the loop area through which external flux passes is small, and any
flux that does couple in produces equal EMF in both conductors that
cancels at a differential receiver.

A representative regulation is the European Union's EMC Directive
(2014/30/EU) and the corresponding harmonised standards in the
EN 61000 and EN 55000 series, which specify radiated and conducted
emission limits, immunity test levels, and the certification
procedure for placing electrical equipment on the European market.
Equivalent regulation in the United States is FCC Part 15 for
unintentional radiators. Compliance is a working constraint on the
design of consumer electronics, industrial equipment, and medical
devices.

\engterm{Lightning} is the failure mode at the high end of natural
electrical activity. A typical cloud-to-ground stroke carries
$30\,\si{\kilo\ampere}$ peak current over a microsecond rise time,
producing $dI/dt \sim 10^{10}\,\si{\ampere\per\second}$ in the
channel and very large radiated fields at distances of hundreds of
metres. The direct strike to a structure is the rare event; the
indirect coupling, in which the magnetic field of the stroke induces
EMF in every loop in nearby equipment, is the common one.
Engineering protection against lightning is a layered discipline:
\begin{itemize}[leftmargin=*]
  \item A \engterm{lightning protection system} provides a preferred
    path for the strike current from the structure's roof to a
    grounding electrode buried at depth, avoiding the interior of the
    structure. International standards IEC 62305 specify the design.
  \item \engterm{Surge protective devices} (SPDs) clamp the voltage on
    incoming power, signal, and antenna lines to a safe value during
    the transient, diverting the surge current to ground. Three SPD
    classes are defined by IEC 61643: Type 1 for direct-strike currents
    at the service entrance, Type 2 for indirect surges in the
    distribution panel, Type 3 for sensitive equipment at the outlet.
  \item \engterm{Bonding} of every grounded structure (panel, conduit,
    rack, equipment chassis) to a common reference at the service
    entrance ensures that during a strike, the entire structure
    floats together rather than allowing potential differences that
    would drive currents through interior loops.
\end{itemize}

A common failure mode of grid-tied photovoltaic installations is
surge damage from indirect lightning: induced EMF in the long DC
cabling between modules and inverter drives currents that exceed the
inverter input ratings, even when no direct strike occurred. The cure
is Type 2 SPDs at the inverter input, properly bonded to the same
grounding system as the array's frame. The same mechanism applies to
wind turbines, where the long blade acts as an antenna for nearby
lightning, and to telecommunications towers, where lightning is a
routine operational hazard rather than an exceptional event.

\begin{principle}[Every cable is an antenna]
Every conductor in an electrical system both radiates and receives
electromagnetic energy at frequencies whose wavelength is comparable
to or larger than the conductor's length. The engineer who treats a
wire as merely a transport for current is correct only when the
wire is short compared to a wavelength; outside that regime, the wire
is an antenna whether the designer intended it to be one or not. The
discipline of EMC, lightning protection, and signal integrity is the
discipline of taking unintended antennas seriously.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Magnetic field around a current-carrying wire]
Build a bench measurement of the magnetic field around a straight
current-carrying wire and verify the $1/r$ law predicted by
Amp\`ere's law.

\textbf{Materials.}
A bench DC power supply (current-limited, $0$ to $5\,\si{\ampere}$ at
$\le 10\,\si{\volt}$); a $1\,\si{\meter}$ straight length of $14$-gauge
copper wire mounted vertically; a Hall-effect magnetometer (a
calibrated probe with a sensitivity of $\sim 10\,\si{\micro\tesla}$ or
better, available as a USB-connected instrument for under \$200, or a
smartphone with a built-in magnetometer and a free measurement app); a
metre rule.

\textbf{Procedure.}
\begin{enumerate}
  \item Calibrate the magnetometer to the local geomagnetic field by
    pointing it in three orthogonal directions and recording the
    components. The Earth's field magnitude at the bench is the
    quadratic sum; verify against a published model (NOAA WMM,
    current as of 2025) within $5\,\%$.
  \item Set the supply to $1.0\,\si{\ampere}$ and measure the
    magnetic-field component perpendicular to the wire at radial
    distances $r = 1, 2, 3, 5, 10, 20, 50\,\si{\centi\meter}$ from
    the wire's centre. Subtract the geomagnetic background.
  \item Repeat at $2.0\,\si{\ampere}$ and $3.0\,\si{\ampere}$.
  \item Plot $B(r)$ on log-log axes; fit a power law $B = A r^{-n}$
    and report $n$ and $A$. Compare to the predicted $\mu_{0}I/(2\pi
    r)$; report the relative error.
\end{enumerate}

\textbf{Report.} The deliverable is a one-page report with the
calibration data, the measured $B(r)$ at each current, the fit
parameters, and a discussion of the dominant error sources. Expected
sources: geomagnetic-background instability ($\sim 1\,\si{\micro
\tesla}$ during the measurement window), probe-position uncertainty
(largest at small $r$), supply-current uncertainty
($\sim 1\,\%$), and finite-wire-length corrections (at $r$ comparable
to or larger than the wire's length, the field falls faster than
$1/r$). The $1/r$ law should be recovered with $n = 1.00\pm 0.05$
across the central two decades of $r$. Failure to recover $n = 1$ at
either end of the range is the project's pedagogic outcome.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Two point charges of $+2\,\si{\micro\coulomb}$ and $-3\,\si{\micro
\coulomb}$ are separated by $30\,\si{\centi\meter}$ in air. Compute
the electric-field magnitude at the midpoint between them, and the
force one would experience by a $1\,\si{\micro\coulomb}$ test charge
placed at the midpoint.

\paragraph{Solution sketch.} The midpoint is $0.15\,\si{\meter}$ from
each charge. Each field magnitude is $E = q/(4\pi\varepsilon_0 r^2) =
(8.99\times 10^9)(q)/(0.15)^2$: the $+2\,\si{\micro\coulomb}$ gives
$E_1 = 7.99\times 10^{5}\,\si{\volt\per\meter}$ pointing away from the
positive charge, the $-3\,\si{\micro\coulomb}$ gives $E_2 = 1.20\times
10^{6}\,\si{\volt\per\meter}$ pointing toward the negative charge. Both
point in the same direction along the line (away from $+$, toward $-$),
so they add: $E = 2.0\times 10^{6}\,\si{\volt\per\meter}$. The force on
the test charge is $F = qE = (10^{-6})(2.0\times 10^{6}) = 2.0\,\si{
\newton}$, directed from the positive toward the negative charge. The
sign trap is to subtract rather than add: along the joining line the
two contributions reinforce.
\end{exercise}

\begin{exercise}
A parallel-plate capacitor has plates of area $50\,\si{\centi\meter
\squared}$ separated by $0.5\,\si{\milli\meter}$ of air. Compute the
capacitance. The plates are charged to $200\,\si{\volt}$; compute the
charge on each plate, the energy stored, and the force of attraction
between the plates.

\paragraph{Solution sketch.} With $A = 50\times 10^{-4}\,\si{\meter
\squared}$ and $d = 5\times 10^{-4}\,\si{\meter}$, $C = \varepsilon_0
A/d = (8.85\times 10^{-12})(5\times 10^{-3})/(5\times 10^{-4}) =
8.85\times 10^{-11}\,\si{\farad} = 88.5\,\si{\pico\farad}$. The charge
is $Q = CV = (8.85\times 10^{-11})(200) = 1.77\times 10^{-8}\,\si{
\coulomb} = 17.7\,\si{\nano\coulomb}$. The stored energy is $U =
\tfrac12 CV^2 = \tfrac12(8.85\times 10^{-11})(200)^2 = 1.77\times
10^{-6}\,\si{\joule}$. The attractive force is $F = \tfrac12
\varepsilon_0 E^2 A$ with $E = V/d = 4\times 10^{5}\,\si{\volt\per
\meter}$, giving $F = \tfrac12(8.85\times 10^{-12})(4\times 10^5)^2
(5\times 10^{-3}) = 3.5\times 10^{-3}\,\si{\newton}$, equivalently
$F = U/d = QE/2$. The factor of one half in the force is the common
error: the field acting on one plate is the field of the \emph{other}
plate, $\sigma/(2\varepsilon_0)$, not the full interior field.
\end{exercise}

\begin{exercise}
A solenoid is wound with $500$ turns over a length of $25\,\si{\centi
\meter}$ on a $2\,\si{\centi\meter}$-radius non-magnetic core. With
$3\,\si{\ampere}$ flowing, compute the interior magnetic field and the
self-inductance.

\paragraph{Solution sketch.} The turn density is $n = 500/0.25 = 2000\,
\si{\per\meter}$. The interior field is $B = \mu_0 n I = (4\pi\times
10^{-7})(2000)(3) = 7.54\times 10^{-3}\,\si{\tesla} = 7.5\,\si{\milli
\tesla}$. The inductance follows from flux linkage: $L = \mu_0 n^2 A
\ell$ with $A = \pi(0.02)^2 = 1.26\times 10^{-3}\,\si{\meter\squared}$
and $\ell = 0.25\,\si{\meter}$, giving $L = (4\pi\times 10^{-7})(2000)^2
(1.26\times 10^{-3})(0.25) = 1.58\times 10^{-3}\,\si{\henry} \approx
1.6\,\si{\milli\henry}$. Equivalently $L = N\Phi/I = N B A/I$ with
$N = 500$. The check is that $\tfrac12 LI^2 = 7.1\,\si{\milli\joule}$
matches $\tfrac12(B^2/\mu_0)\times(\text{interior volume})$.
\end{exercise}

\begin{exercise}
A coaxial cable has inner conductor of radius $0.4\,\si{\milli\meter}$
and outer conductor of inner radius $2.0\,\si{\milli\meter}$, with
polyethylene dielectric ($\varepsilon_{r} = 2.25$, $\mu_{r} = 1$).
Compute the capacitance per metre, the inductance per metre, and the
characteristic impedance.

\paragraph{Solution sketch.} The ratio is $b/a = 2.0/0.4 = 5$, so
$\ln(b/a) = 1.609$. Capacitance per metre: $C' = 2\pi\varepsilon_0
\varepsilon_r/\ln(b/a) = 2\pi(8.85\times 10^{-12})(2.25)/1.609 =
7.78\times 10^{-11}\,\si{\farad\per\meter} = 77.8\,\si{\pico\farad\per
\meter}$. Inductance per metre: $L' = (\mu_0/2\pi)\ln(b/a) = (2\times
10^{-7})(1.609) = 3.22\times 10^{-7}\,\si{\henry\per\meter} = 322\,\si{
\nano\henry\per\meter}$. Characteristic impedance: $Z_0 = \sqrt{L'/C'}
= \sqrt{3.22\times 10^{-7}/7.78\times 10^{-11}} = 64\,\si{\ohm}$;
equivalently $Z_0 = (60/\sqrt{\varepsilon_r})\ln(b/a) = (60/1.5)(1.609)
= 64\,\si{\ohm}$. The script \repopath{code/coax\_parameters.py}
reproduces these values.
\end{exercise}

\begin{exercise}
A circular loop of radius $5\,\si{\centi\meter}$ carries a current that
ramps linearly from $0$ to $2\,\si{\ampere}$ in $10\,\si{\milli\second}$.
Compute the rate of change of the magnetic flux through a concentric
loop of radius $1\,\si{\centi\meter}$, and the EMF induced around the
inner loop (assume the inner loop is in the central region where the
outer loop's field is approximately uniform and given by $B =
\mu_{0}I/(2R)$).

\paragraph{Solution sketch.} The outer loop's centre field is $B =
\mu_0 I/(2R)$ with $R = 0.05\,\si{\meter}$. The flux through the inner
loop of radius $r = 0.01\,\si{\meter}$, area $a = \pi r^2 = 3.14\times
10^{-4}\,\si{\meter\squared}$, is $\Phi = Ba = \mu_0 I a/(2R)$. Its
rate of change tracks $\dd I/\dd t = 2/(10\times 10^{-3}) = 200\,\si{
\ampere\per\second}$: $\dd\Phi/\dd t = \mu_0 a/(2R)\cdot\dd I/\dd t =
(4\pi\times 10^{-7})(3.14\times 10^{-4})/(0.1)\times 200 = 7.9\times
10^{-7}\,\si{\weber\per\second}$. The induced EMF around the single
inner loop is $\mathcal{E} = -\dd\Phi/\dd t = -0.79\,\si{\micro\volt}$,
a magnitude of about $0.8\,\si{\micro\volt}$. The smallness of the
number is the lesson: tightly coupled single turns at modest $\dd I/
\dd t$ induce microvolts, which is why transformers use many turns and
high frequencies.
\end{exercise}

\begin{exercise}
The solar irradiance at Earth's surface on a clear day is around
$1000\,\si{\watt\per\meter\squared}$. Compute the peak electric-field
amplitude and the peak magnetic-field amplitude of the plane-wave
component.

\paragraph{Solution sketch.} From $\langle S\rangle = \tfrac12
\varepsilon_0 c E_0^2$, solve for the peak electric field: $E_0 =
\sqrt{2\langle S\rangle/(\varepsilon_0 c)} = \sqrt{2(1000)/((8.85\times
10^{-12})(3\times 10^8))} = 868\,\si{\volt\per\meter}$. The peak
magnetic field is $B_0 = E_0/c = 868/(3\times 10^8) = 2.9\times
10^{-6}\,\si{\tesla} = 2.9\,\si{\micro\tesla}$, comparable in magnitude
to the Earth's static field but oscillating at optical frequency. The
script \repopath{code/plane\_wave.py} returns the same values and the
$1361\,\si{\watt\per\meter\squared}$ space-side case.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Starting from $\oint \vect{E}\cdot \dd\vect{A} = Q_{\text{enc}}/
\varepsilon_{0}$, derive Coulomb's law for the field of a point
charge. State the symmetry assumption that closes the integral.

\paragraph{Solution sketch.} The symmetry assumption is that a point
charge at the origin produces a field that is purely radial and depends
only on $r$, so that on a sphere of radius $r$ centred on the charge,
$\vect{E}\cdot\dd\vect{A} = E(r)\,\dd A$ with $E(r)$ constant over the
sphere. The flux integral then factors: $\oint E(r)\,\dd A = E(r)\cdot
4\pi r^2$. Setting this equal to $q/\varepsilon_0$ gives $E(r) =
q/(4\pi\varepsilon_0 r^2)$, and $\vect{F} = q'\vect{E}$ on a second
charge recovers Coulomb's law $F = q q'/(4\pi\varepsilon_0 r^2)$. The
derivation uses no integration beyond the area of a sphere; the entire
content is the symmetry argument that makes $E$ constant on the chosen
surface.
\end{exercise}

\begin{exercise}
Derive the displacement current term in the Amp\`ere-Maxwell law by
demanding consistency between Amp\`ere's law without displacement
current and the continuity equation $\partial \rho/\partial t +
\divg \vect{J} = 0$. State precisely where the bare Amp\`ere law
fails.

\paragraph{Solution sketch.} The bare law is $\curl\vect{B} = \mu_0
\vect{J}$. Take the divergence of both sides: $\divg(\curl\vect{B}) =
\mu_0\divg\vect{J}$. The left side is identically zero, since the
divergence of any curl vanishes. So the bare law forces $\divg\vect{J}
= 0$ everywhere and always. Continuity says $\divg\vect{J} = -\partial
\rho/\partial t$, which is non-zero wherever charge accumulates, as on a
charging capacitor plate. The contradiction is exactly there: the bare
law fails whenever $\partial\rho/\partial t \neq 0$. Adding a term
$\mu_0\varepsilon_0\,\partial\vect{E}/\partial t$ and using Gauss's law
$\divg\vect{E} = \rho/\varepsilon_0$ gives $\divg(\mu_0\vect{J} +
\mu_0\varepsilon_0\,\partial\vect{E}/\partial t) = \mu_0(-\partial\rho/
\partial t) + \mu_0\,\partial\rho/\partial t = 0$ identically. The
divergence is now consistent, and the added term is the displacement
current.
\end{exercise}

\begin{exercise}
Derive the Poynting theorem $\partial (u_{e}+u_{m})/\partial t +
\divg \vect{S} = -\vect{J}\cdot\vect{E}$ by combining $\vect{E}\cdot
(\curl \vect{B} = \mu_{0}\vect{J} + \mu_{0}\varepsilon_{0}\partial
\vect{E}/\partial t)$ with $\vect{B}\cdot(\curl \vect{E} = -\partial
\vect{B}/\partial t)$ and using the vector identity $\divg(\vect{E}
\times \vect{B}) = \vect{B}\cdot \curl \vect{E} - \vect{E}\cdot \curl
\vect{B}$.

\paragraph{Solution sketch.} Dot the Amp\`ere-Maxwell law with $\vect{E}$:
$\vect{E}\cdot\curl\vect{B} = \mu_0\vect{J}\cdot\vect{E} + \mu_0
\varepsilon_0\vect{E}\cdot\partial\vect{E}/\partial t$. Dot Faraday's
law with $\vect{B}$: $\vect{B}\cdot\curl\vect{E} = -\vect{B}\cdot
\partial\vect{B}/\partial t$. Subtract the first from the second and
apply the identity, $\divg(\vect{E}\times\vect{B}) = \vect{B}\cdot
\curl\vect{E} - \vect{E}\cdot\curl\vect{B}$, on the left. Recognise the
two time-derivative products as $\vect{E}\cdot\partial\vect{E}/\partial
t = \tfrac12\partial|\vect{E}|^2/\partial t$ and likewise for $\vect{B}$.
Divide through by $\mu_0$ and collect: $\partial/\partial t(\tfrac12
\varepsilon_0|\vect{E}|^2 + \tfrac{1}{2\mu_0}|\vect{B}|^2) + \divg(
\vect{E}\times\vect{B}/\mu_0) = -\vect{J}\cdot\vect{E}$, which is the
Poynting theorem with $\vect{S} = \vect{E}\times\vect{B}/\mu_0$. The
right side is the volumetric work the field does on charge, the resistive
loss $\vect{J}\cdot\vect{E}$.
\end{exercise}

\begin{exercise}
Derive the four boundary conditions at the interface between two
linear isotropic media by applying Maxwell's equations in integral
form to a pillbox (for normal components) and to a rectangular loop
(for tangential components) straddling the interface, in the limit
that the pillbox height and the loop's normal dimension go to zero.

\paragraph{Solution sketch.} Normal components: apply $\oint\vect{D}\cdot
\dd\vect{A} = Q_{f,\text{enc}}$ to a flat pillbox of face area $\dd A$
and vanishing height. As the height goes to zero the side-wall flux
vanishes and only the two faces contribute, giving $(D_{2n} - D_{1n})
\,\dd A = \sigma_f\,\dd A$, so $D_{2n} - D_{1n} = \sigma_f$. The same
pillbox with $\divg\vect{B}=0$ gives $B_{2n} - B_{1n} = 0$. Tangential
components: apply $\oint\vect{E}\cdot\dd\vect{\ell} = -\dd\Phi_B/\dd t$
to a thin rectangular loop straddling the interface with long sides
$\dd\ell$ tangential and short sides normal and vanishing. As the loop
area goes to zero the flux term vanishes, leaving $(\vect{E}_{2t} -
\vect{E}_{1t})\,\dd\ell = 0$, so tangential $\vect{E}$ is continuous.
The same loop with the Amp\`ere-Maxwell law retains the enclosed free
surface current $\vect{K}_f$ as the area shrinks, giving $\unitvec{n}
\times(\vect{H}_2 - \vect{H}_1) = \vect{K}_f$. The trick in each case is
that one dimension of the Gaussian region is sent to zero, killing the
volume or area term and leaving only the surface source.
\end{exercise}

\begin{exercise}
Show that the electromagnetic field's momentum density is
$\vect{g} = \varepsilon_{0}\vect{E}\times \vect{B} = \vect{S}/c^{2}$,
and that radiation pressure on a perfectly absorbing surface is
$P = I/c$ where $I$ is the time-averaged intensity. Compute the
radiation pressure of solar irradiance at $1361\,\si{\watt\per\meter
\squared}$.

\paragraph{Solution sketch.} The field momentum density is the Poynting
vector divided by $c^2$: $\vect{g} = \vect{S}/c^2 = \varepsilon_0
\vect{E}\times\vect{B}$, using $\vect{S} = \vect{E}\times\vect{B}/\mu_0$
and $1/(\mu_0 c^2) = \varepsilon_0$. The momentum delivered per unit
time per unit area to an absorbing surface is the momentum flux, which
equals the energy flux divided by $c$ for light, so the pressure is
$P = \langle S\rangle/c = I/c$. (A perfect reflector doubles this.)
Numerically $P = 1361/(3\times 10^8) = 4.5\times 10^{-6}\,\si{\pascal}$,
about $4.5\,\si{\micro\pascal}$: tiny on a hand, but the basis of solar
sails and a real perturbation on spacecraft attitude over long
missions. The script \repopath{code/plane\_wave.py} reports the same
pressure.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the energy stored in the magnetic field of the Earth in the
volume from the surface to $1$ Earth radius altitude, treating the
field as roughly dipolar with surface magnitude $50\,\si{\micro\tesla}$.
State assumptions about the field's variation with altitude.

\paragraph{Solution sketch.} Energy density is $u_m = B^2/(2\mu_0)$. At the
surface, $u_m = (50\times 10^{-6})^2/(2\times 4\pi\times 10^{-7}) =
1.0\times 10^{-3}\,\si{\joule\per\meter\cubed}$. A dipole field falls as
$1/r^3$, so $u_m$ falls as $1/r^6$; most of the energy in the shell sits
near the surface. Take the shell volume from $R_E = 6.4\times 10^6\,\si{
\meter}$ to $2R_E$ as roughly $V \sim 4\pi R_E^2\cdot R_E = 4\pi R_E^3
\approx 3\times 10^{21}\,\si{\meter\cubed}$, but weight by the $1/r^6$
falloff, which cuts the effective volume to a fraction of order $0.3$ of
a one-radius shell. An order-of-magnitude estimate is $U \sim u_m\cdot
0.3\cdot 4\pi R_E^3/3 \sim (10^{-3})(0.3)(1.1\times 10^{21}) \sim
3\times 10^{17}\,\si{\joule}$. The dominant assumption is the dipole
$1/r^3$ law, which the real magnetosphere violates above a few Earth
radii where the solar wind distorts it; the estimate is good to a factor
of a few, no better.
\end{exercise}

\begin{exercise}
Estimate the peak voltage induced across the terminals of a
$1\,\si{\meter\squared}$ open loop placed $100\,\si{\meter}$ from a
lightning channel carrying a peak current of $30\,\si{\kilo\ampere}$
with rise time $1\,\si{\micro\second}$. State the dominant
uncertainty in the answer.

\paragraph{Solution sketch.} The channel field at $r = 100\,\si{\meter}$
is $B = \mu_0 I/(2\pi r) = (4\pi\times 10^{-7})(30\times 10^3)/(2\pi
\times 100) = 6.0\times 10^{-5}\,\si{\tesla}$ at peak. The field rises in
$\tau = 1\,\si{\micro\second}$, so $\dd B/\dd t \approx B/\tau = 6.0
\times 10^{-5}/10^{-6} = 60\,\si{\tesla\per\second}$. The EMF in a
$1\,\si{\meter\squared}$ loop is $\mathcal{E} = A\,\dd B/\dd t = 60\,\si{
\volt}$, of order tens of volts. The dominant uncertainty is the loop's
orientation and effective area relative to the field, which can swing the
flux linkage by a full order of magnitude depending on geometry; the
$1/r$ near-field approximation and the assumed linear current ramp add
smaller errors. The lesson is that a modest loop a hundred metres from a
strike sees tens of volts with no direct contact, which is why induced
surges, not direct strikes, are the common failure.
\end{exercise}

\begin{exercise}
Estimate the time it would take for the field of a $1\,\si{\farad}$
capacitor charged to $100\,\si{\volt}$ to discharge through the
leakage resistance of typical air at $50\,\%$ relative humidity. State
the assumptions and report a factor-of-ten confidence range.

\paragraph{Solution sketch.} The discharge time constant is $\tau = RC$.
The capacitance is $C = 1\,\si{\farad}$. The leakage resistance depends
on the geometry and the path through air; take a representative bench
gap of $1\,\si{\centi\meter}$ over $1\,\si{\centi\meter\squared}$ of
plate and air conductivity in the range $\sigma \sim 10^{-14}$ to
$10^{-13}\,\si{\siemens\per\meter}$ at moderate humidity. Then $R =
\ell/(\sigma A) = 10^{-2}/(10^{-13}\times 10^{-4}) \sim 10^{15}\,\si{
\ohm}$. The time constant is $\tau = RC \sim 10^{15}\,\si{\second}$,
which is enormous, longer than the age of the universe, so air leakage
across a clean gap is not what discharges a real capacitor. The
assumption that fails is that air across the gap is the path; in
practice surface leakage across the dielectric, the dielectric's own
bulk conductivity, and contamination dominate, dropping the practical
time constant to days or hours. The factor-of-ten range on the air-only
estimate spans $10^{14}$ to $10^{16}\,\si{\second}$, and the point of
the exercise is that the idealised number is absurd: the real leakage
path is never clean air.
\end{exercise}

\begin{exercise}
Estimate the radio-frequency energy radiated by a $100\,\si{\watt}$
incandescent bulb at the bulb's filament-fluctuation frequencies.
State the bandwidth assumed and the radiation efficiency at those
frequencies.

\paragraph{Solution sketch.} An incandescent filament runs near constant
temperature; its electrical fluctuations are dominated by the
power-line frequency and its low harmonics ($50$ or $60\,\si{\hertz}$
and a few multiples) plus thermal noise. Take a fluctuation bandwidth
of order $1\,\si{\kilo\hertz}$ and a fluctuating power that is a small
fraction, say $1\,\%$, of the $100\,\si{\watt}$, so $\sim 1\,\si{\watt}$
of fluctuating electrical power. The filament is electrically tiny
compared to the wavelength at these frequencies ($\lambda \sim
6000\,\si{\kilo\meter}$ at $50\,\si{\hertz}$), and the radiation
efficiency of an antenna of length $\ell \ll \lambda$ scales as
$(\ell/\lambda)^2$, which is of order $(0.01/6\times 10^6)^2 \sim
10^{-18}$. The radiated RF power is then $\sim 1\,\si{\watt}\times
10^{-18} \sim 10^{-18}\,\si{\watt}$: negligible. The lesson is that
low-frequency sources radiate almost nothing because they are tiny
compared to a wavelength; the EMI problem lives at high frequency,
where the same conductor becomes a fair fraction of a wavelength.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a finite-difference solver for Laplace's equation $\laplacian
\phi = 0$ in two dimensions on a square grid, with Dirichlet boundary
conditions corresponding to a parallel-plate capacitor with finite-
size plates. Compare the computed capacitance against the
infinite-plate formula and identify the fringing-field correction as
a function of plate aspect ratio.

\paragraph{Solution sketch.} Discretise $\phi$ on an $N\times N$ grid and
relax the five-point stencil $\phi_{ij} = \tfrac14(\phi_{i+1,j} +
\phi_{i-1,j} + \phi_{i,j+1} + \phi_{i,j-1})$ at every free node, holding
the two plate rows at $\pm V$ and the box boundary at zero, iterating
until the maximum change per sweep falls below tolerance. Recover the
capacitance from the stored energy, $C = 2U/V^2$ with $U = \tfrac12
\varepsilon_0\sum|\grad\phi|^2\,\dd A$. The computed $C$ exceeds the
infinite-plate $\varepsilon_0 A/d$ because the fringing field stores
extra energy beyond the plate edges; the surplus is a larger fraction
of the total at small aspect ratio (plate width over gap) and shrinks
toward zero as the plates grow wide. The reference implementation is
\repopath{code/laplace\_capacitor.py}, which prints $C/C_{\infty}$ for
several aspect ratios and converges in a few thousand Gauss-Seidel
sweeps. The expected result is a fringing surplus of tens of percent at
aspect ratio of a few, falling as the plates widen.
\end{exercise}

\begin{exercise}
Implement a numerical evaluation of the Biot-Savart integral for a
finite straight wire segment, and compare against the analytical
$B(r,\theta) = (\mu_{0} I/4\pi r)(\sin\theta_{2}-\sin\theta_{1})$
result. Use the simulation to map the field around a square current
loop and identify the centre-of-loop value.

\paragraph{Solution sketch.} Discretise the wire into many short elements
$\dd\vect{\ell}$ and sum $\dd\vect{B} = (\mu_0 I/4\pi)\,\dd\vect{\ell}
\times\unitvec{r}/r^2$ over them with midpoint quadrature. For a single
finite segment seen perpendicular at distance $r$, the numerical sum
must match $(\mu_0 I/4\pi r)(\sin\theta_2 - \sin\theta_1)$, and it
approaches the infinite-wire $\mu_0 I/(2\pi r)$ as the segment
lengthens. For a square loop of side $a$, sum the four sides; the centre
value should converge to $2\sqrt{2}\,\mu_0 I/(\pi a)$, which is four
times the contribution of one side evaluated at the loop centre. The
reference implementation \repopath{code/biot\_savart\_segment.py}
reproduces both checks; the convergence rate is set by the number of
quadrature points and is poorest where the field point is closest to the
wire, since the integrand peaks there.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a single-layer air-core solenoid that produces $10\,\si{\milli
\tesla}$ on its axis with no more than $20\,\si{\watt}$ dissipation.
Specify the wire gauge, number of turns, length, radius, and supply
current. State the dominant thermal-management constraint and the
mass of copper required.

\paragraph{Solution sketch.} Target $B = 10\,\si{\milli\tesla} = \mu_0 n
I$, so $nI = B/\mu_0 = 10^{-2}/(4\pi\times 10^{-7}) = 7.96\times 10^3\,
\si{\ampere\per\meter}$ (amp-turns per metre). Pick a workable
operating point: $I = 4\,\si{\ampere}$ needs $n \approx 2000\,\si{\per
\meter}$, which suits a single layer of about $0.5\,\si{\milli\meter}$
enamelled wire (AWG 24, roughly $0.51\,\si{\milli\meter}$ overall).
Choose length $\ell = 0.2\,\si{\meter}$ (so $N = n\ell = 400$ turns) and
radius $a = 2\,\si{\centi\meter}$. The wire length is $L_w = N\cdot 2\pi
a = 400\times 0.126 = 50\,\si{\meter}$; AWG 24 has resistance about
$0.084\,\si{\ohm\per\meter}$, so $R \approx 4.2\,\si{\ohm}$ and the
dissipation is $I^2 R = 16\times 4.2 = 67\,\si{\watt}$, over the
$20\,\si{\watt}$ budget. Bring it under budget by raising the current and
using a thicker conductor with fewer turns: for a single-layer winding the
dissipation at fixed field scales as $n$, so going to $I = 16\,\si{
\ampere}$ with $n = 500\,\si{\per\meter}$ (about a $2\,\si{\milli\meter}$
conductor) gives $I^2 R \approx 17\,\si{
\watt}$ for the same field. The dominant constraint is heat removal from
a long thin winding with poor surface-to-volume ratio; natural
convection limits the surface flux to a few hundred $\si{\watt\per\meter
\squared}$, so the winding surface must exceed roughly $0.1\,\si{\meter
\squared}$ or forced cooling is needed. The copper mass is $\rho_{Cu}
\times L_w\times A_{wire} = 8960\times 50\times (2\times 10^{-7})
\approx 0.09\,\si{\kilogram}$.
\end{exercise}

\begin{exercise}
Design a single-point ground for a small audio-recording studio
containing a microphone preamp, a computer audio interface, and a
monitor amplifier. Sketch the grounding topology, identify potential
ground loops, and specify the bonding cable gauge.

\paragraph{Solution sketch.} Bring every chassis earth to one physical node
(a star point, typically the audio interface or a dedicated bus bar)
with separate radial conductors, so no two devices share a return path
that could enclose loop area. The ground loops to watch are: the preamp
and interface both earthed through their own mains plugs while also
joined by the audio cable shield (cut the shield at one end, or use a
balanced connection); the computer earthed through its own supply and
through USB to the interface; the monitor amp earthed through its mains
plug and through the speaker-level or line return. Use balanced
(differential) interconnects between devices wherever the equipment
supports them, so common-mode hum is rejected at the receiver. The
bonding conductors carry no signal current, only fault and equalising
current, so a modest gauge suffices for the bonds themselves; a common
choice is $2.5\,\si{\milli\meter\squared}$ (about AWG 14) for the star
bonds, sized for fault current and mechanical robustness rather than
for the milliampere signal currents. The deliverable is a star diagram
with one node and radial bonds, plus the note that the safety earth on
each mains plug must not be defeated to break a loop.
\end{exercise}

\begin{exercise}
Design a Type 2 surge protective device specification for a $5\,\si{
\kilo\watt}$ residential photovoltaic inverter's DC input. Specify
the maximum continuous operating voltage, nominal discharge current,
voltage protection level, and response time. Reference IEC 61643-31
or the equivalent national standard.

\paragraph{Solution sketch.} A $5\,\si{\kilo\watt}$ residential array
typically runs at a few hundred volts DC; size the SPD's maximum
continuous operating voltage ($U_{cp}$) above the array's open-circuit
voltage at the coldest expected temperature with margin, for example
$U_{cp} \geq 1000\,\si{\volt}$ DC for a string whose cold open-circuit
voltage reaches the high $600$s to low $800$s of volts. Choose a Type 2
device (indirect surges) with nominal discharge current $I_n$ of
$20\,\si{\kilo\ampere}$ ($8/20\,\si{\micro\second}$ waveform) and a
maximum discharge current $I_{max}$ of $40\,\si{\kilo\ampere}$. The
voltage protection level $U_p$ should sit comfortably below the
inverter's DC-input surge withstand, typically $U_p \leq 4\,\si{\kilo
\volt}$, and lower if the inverter datasheet demands it. Response time
for a metal-oxide-varistor device is under $25\,\si{\nano\second}$. The
device must be bonded with a short, low-inductance lead to the same
grounding electrode as the array frame, because lead inductance adds
$L\,\dd I/\dd t$ to $U_p$ and can dominate at the steep surge fronts.
The governing standard is IEC 61643-31 for photovoltaic SPDs (with the
installation requirements in IEC 60364-7-712).
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{standard}

\begin{exercise}
A magnetic-field reading at the bench shows $300\,\si{\micro\tesla}$
at $20\,\si{\centi\meter}$ from a thick power cable. Identify the
current flowing in the cable, and identify whether the cable carries
DC or AC by what additional measurement would distinguish them.

\paragraph{Solution sketch.} Invert $B = \mu_0 I/(2\pi r)$: $I = 2\pi r
B/\mu_0 = 2\pi(0.2)(300\times 10^{-6})/(4\pi\times 10^{-7}) = 300\,\si{
\ampere}$. The DC-versus-AC question is answered by the probe's
response over time: a DC current gives a steady reading on a Hall probe
and zero reading on a clamp meter or search coil that responds only to
$\dd B/\dd t$; an AC current gives a reading that oscillates at the line
frequency and is detected by an inductive pickup. So the distinguishing
measurement is either a search-coil pickup (responds to AC only) or
watching the Hall probe for a steady versus oscillating value. Note the
clamp around a two-conductor cable carrying balanced go-and-return
current would read near zero; the $300\,\si{\micro\tesla}$ reading
implies a net unbalanced current, such as a single conductor or a fault.
\end{exercise}

\begin{exercise}
A capacitor stamped ``$10\,\si{\micro\farad}, 25\,\si{\volt}$'' is
measured at $9.7\,\si{\micro\farad}$ on an LCR meter at $1\,\si{\kilo
\hertz}$ and $8.4\,\si{\micro\farad}$ at $100\,\si{\kilo\hertz}$.
Identify the most likely cause of the frequency dependence, and the
secondary measurement that would distinguish dielectric absorption
from equivalent series resistance.

\paragraph{Solution sketch.} The apparent capacitance falling with
frequency points to the device's parasitics: a real capacitor is a
series $R$-$L$-$C$ network, and the equivalent series inductance plus
the dielectric's frequency-dependent permittivity both reduce the
measured value as frequency rises. For an electrolytic at $100\,\si{
\kilo\hertz}$ the most likely cause is the combination of equivalent
series resistance (ESR) and inductance shifting the impedance away from
ideal $1/(\omega C)$, together with the polar dielectric's relaxation.
To separate dielectric absorption from ESR, measure the loss tangent or
dissipation factor versus frequency: ESR produces a loss that rises with
frequency as $\omega C \cdot \text{ESR}$, while dielectric relaxation
produces a loss peak at the relaxation frequency. A complementary test
is a dielectric-absorption ("soakage") measurement: charge, briefly
discharge, then watch the voltage recover; the recovery isolates the
slow dielectric mechanism from the resistive ESR.
\end{exercise}

\begin{exercise}
A coaxial cable in a test fixture shows an unexpectedly high VSWR.
Identify three candidate causes (connector mismatch, dielectric
damage, length-and-cut frequency interaction), and the
time-domain-reflectometry signature that distinguishes them.

\paragraph{Solution sketch.} A high voltage standing-wave ratio means
power is being reflected by an impedance discontinuity somewhere along
the line. Time-domain reflectometry (TDR) sends a fast step and watches
the echo; the position of the reflection in time, multiplied by the
propagation velocity, locates the fault. A connector mismatch shows a
sharp localised reflection at the connector position, whose sign tells
whether the discontinuity is high or low impedance. Dielectric damage
(crush, water ingress) shows a reflection partway along the cable at the
damage point, often broadened and with a velocity change beyond it.
A length-and-cut frequency interaction is not a localised fault at all:
the reflection comes from the far-end load, and the VSWR varies
periodically with frequency as the round-trip electrical length passes
through half-wavelengths, with no internal TDR echo. So the
distinguishing signature is the time position and shape of the echo:
at a connector, partway along the dielectric, or only at the far end.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A laboratory data-acquisition system shows $60\,\si{\hertz}$ hum on
every channel even when the input is shorted at the connector.
Identify the failure mode and three measurements that would
distinguish a ground-loop cause from a power-supply-coupling cause.

\paragraph{Solution sketch.} Hum present on every channel with the input
shorted means the interference is entering after the input, through the
shared ground or the power supply rather than through the signal path.
The most common cause is a ground loop: two separate earth paths
enclosing area through which mains-frequency flux induces an EMF.
Three discriminating measurements: (1) lift the chassis earth through an
isolation transformer (safely, briefly) and watch whether the hum drops
sharply, which implicates a ground loop; (2) measure the hum frequency
content, where a pure $60\,\si{\hertz}$ tone suggests inductive
ground-loop pickup while strong harmonics ($120\,\si{\hertz}$ and up)
suggest rectifier ripple from the power supply; (3) probe the supply
rails directly for ripple at $120\,\si{\hertz}$, which appears in
supply-coupling but not in a clean ground loop. The cure for the loop is
a single-point ground; the cure for supply coupling is better filtering
and regulation.
\end{exercise}

\begin{exercise}
A photovoltaic inverter fails after a thunderstorm with no direct
strike to the array. Identify the failure mechanism, the design
defect that allowed it, and the IEC standard that addresses it.

\paragraph{Solution sketch.} The mechanism is induced surge: a nearby
lightning stroke produces a fast-changing magnetic field that links the
loop area formed by the long DC cabling between modules and inverter,
inducing an EMF (Faraday's law) that drives a current exceeding the
inverter's input surge rating, even with no direct strike. The design
defect is the absence of surge protection at the inverter DC input
combined with a large, unmanaged DC-cable loop area and inadequate
bonding of the array frame to the inverter grounding. The cure is a
Type 2 SPD at the inverter DC input, bonded to the same grounding
electrode as the frame, plus routing the DC positive and negative
conductors close together to shrink the loop area. The governing
standards are IEC 61643-31 (photovoltaic SPDs) and the lightning
protection framework IEC 62305; this chapter narrates the mechanism only
and names no specific accident.
\end{exercise}

\begin{exercise}
An MRI suite reports inconsistent image quality correlated with the
time of day. Identify two electromagnetic causes (DC-traction-power
disturbance from a nearby tram line, lift-motor commutation during
business hours) and the measurement that distinguishes them.

\paragraph{Solution sketch.} MRI imaging is acutely sensitive to small
fluctuations in the static field and to stray AC fields in the imaging
band. A nearby tram or light-rail line carries large DC traction
currents whose return paths shift through the ground as trains move,
producing slow, large-amplitude DC field excursions that track the
transit schedule. Lift (elevator) motors commutate when cars run during
business hours, producing AC and broadband disturbances correlated with
occupancy. The distinguishing measurement is a continuous fluxgate (DC-
capable) magnetometer log in the scanner room: the tram signature is a
slow, low-frequency DC drift correlated with the transit timetable,
while the lift signature is intermittent higher-frequency bursts
correlated with car movement. Spectral analysis of the log separates the
near-DC drift from the motor-commutation bursts, and correlating each
against the tram timetable and the lift call log identifies the source.
\end{exercise}

\begin{exercise}
A radio receiver's sensitivity drops by $20\,\si{\decibel}$ when a
nearby LED desk lamp is switched on. Identify the failure mode, the
regulatory limit that should have prevented it, and the
electromagnetic-emissions test that the lamp manufacturer should have
performed.

\paragraph{Solution sketch.} The failure mode is conducted and radiated
EMI from the lamp's switching driver: a cheap LED supply switches at
tens to hundreds of kilohertz with fast edges whose harmonics extend
into the radio bands, raising the receiver's noise floor and cutting its
sensitivity. The regulatory limit that should have prevented it is the
conducted- and radiated-emission limit for lighting equipment, in the
EU the harmonised standard EN 55015 (CISPR 15) under the EMC Directive
2014/30/EU, and in the United States FCC Part 15 for unintentional
radiators. The test the manufacturer should have run is a CISPR 15
conducted-emissions scan on the mains terminals plus a radiated-emission
scan over the relevant frequency range in a calibrated setup (line
impedance stabilisation network for conducted, an antenna at a fixed
distance for radiated). A compliant driver uses input filtering and
edge-rate control to keep harmonics under the limit line. The chapter
narrates the mechanism and the standards; it names no specific product.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer proposes to route a high-current power cable and a
low-level signal cable through the same conduit ``to save trays.''
Identify the EMC argument against the proposal and the conditions
under which the practice would nevertheless be acceptable.

\paragraph{Solution sketch.} The argument against is crosstalk: the
high-current cable's changing magnetic field couples inductively into
the signal cable's loop, and its changing voltage couples capacitively,
injecting interference proportional to the source's $\dd I/\dd t$ and
$\dd V/\dd t$ and to the run length over which the two share the
conduit. A low-level signal is the worst victim because the coupled
noise is large relative to the wanted signal. The practice becomes
acceptable when the coupling is made negligible: the signal is balanced
(differential) and shielded so common-mode pickup is rejected; the
signal is twisted pair so the loop area is tiny; the power cable carries
DC or low, slowly varying current with small $\dd I/\dd t$; physical
separation and a grounded barrier inside the conduit are maintained; and
the run is short. Many cabling codes set a minimum separation between
power and signal that, if met, permits shared routing. The honest answer
to the junior engineer is conditional, not a flat refusal.
\end{exercise}

\begin{exercise}
A consulting client asks whether they should use a single large
lightning rod at the apex of a $30\,\si{\meter}$ tall industrial
building or a grid of smaller rods distributed across the roof.
Identify the IEC 62305 rolling-sphere model and explain which
approach the model favours.

\paragraph{Solution sketch.} The rolling-sphere method of IEC 62305 models
the lightning attachment point geometrically: a sphere of a defined
radius (set by the protection level, for example $20$ to $60\,\si{
\meter}$) is rolled over the structure, and any surface the sphere
touches is a possible strike point, while any region the sphere cannot
reach is protected. A single rod at the apex of a $30\,\si{\meter}$
building leaves much of the roof exposed, because the sphere rolling
down from the rod still touches the roof edges and any equipment set
back from the apex. A distributed grid of rods, sized and spaced so the
sphere cannot dip to the roof surface between them, protects the whole
roof. The model favours the distributed grid for a large flat or
equipment-laden roof; the single tall rod protects only a cone-like zone
beneath it and is adequate only for a small footprint. The judgment is
to apply the rolling-sphere geometry to the actual roof plan rather than
to argue from the height of a single rod.
\end{exercise}

\begin{exercise}
A regulator is considering raising EMC-emission limits for consumer
electronics to support 5G mobile-radio rollout. Identify the trade-
off between consumer-product cost and radio-service protection, and
the working compromise that has been adopted internationally.

\paragraph{Solution sketch.} Raising emission limits lets manufacturers
spend less on filtering and shielding, lowering product cost, at the
price of a higher ambient noise floor that degrades sensitive radio
services such as mobile reception. Tightening limits protects radio at
the cost of more expensive products. The working compromise adopted
internationally is frequency-dependent, banded limit lines: emission
limits are set per frequency band, stricter in the bands allocated to
protected radio services and looser elsewhere, with separate conducted
and radiated limits and defined measurement methods (the CISPR family
of standards). The regulator does not pick a single number; it shapes
the limit curve to protect the spectrum where protection matters while
relaxing it where no sensitive service operates. The judgment is to
frame the question as a per-band limit curve, not a single global
threshold.
\end{exercise}

\begin{exercise}
An industrial facility experiences periodic motor-drive failures
attributed to ``power quality.'' A consultant proposes installing
$\$50\,000$ of harmonic filters; another proposes $\$5\,000$ of
isolation transformers. Identify what measurements would settle the
choice and the conditions under which each remedy is appropriate.

\paragraph{Solution sketch.} The measurement that settles it is a power-
quality survey at the drive's supply: log the voltage and current
waveforms and compute the harmonic spectrum (total harmonic distortion
and the individual harmonic magnitudes), the flicker, and any transient
events, over a representative window that captures the failure pattern.
If the dominant problem is harmonic current distortion (large fifth,
seventh, eleventh harmonics from nonlinear loads), harmonic filters are
the appropriate remedy and the $\$50{,}000$ spend may be justified. If
the problem is common-mode noise, ground-loop coupling, or a need to
break a galvanic path, an isolation transformer at $\$5{,}000$ solves it
and the expensive filters would not. If the failures correlate with
voltage sags or transients from other equipment, neither remedy fits and
the cure is a line conditioner or upstream fault correction. The honest
engineering answer is to measure before specifying: the spectrum and the
event log tell which mechanism is operating, and each remedy treats only
its own mechanism.
\end{exercise}
