\chapter{Healthcare systems}
\label{vol12:ch06}

\epigraph{Healthcare is engineering: it is the application of disciplined intervention to human bodies under constraint, failure, scale, and responsibility.}{Author}

\chapmeta{Half-life: medium-life. Archetypes: control, scaling, ethics. Exercise target: 20. Project track: simulation (assess one healthcare system's resilience). Hazard class: high (patient-safety; pandemic-scale stakes). Reader path default: standard, with the system and public-health sections core and the pandemic-engineering section core. Named cases: COVID-19 ventilator and PPE supply chain 2020 (\cite{paper:cohen-pandemic-supply-2020}); Mid Staffordshire NHS Foundation Trust 2005-2009 (\cite{acc:francis-mid-staffs-2013}); Therac-25 (\cite{paper:leveson-turner1993}, registered).}

\noindent
Healthcare is engineered at every layer at which it operates. The hospital is an engineered facility with utilities, life-safety systems, infection-control airflows, and medical-device infrastructure. The medical device is an engineered artifact with safety, reliability, and regulatory requirements. The pharmaceutical is a manufactured chemical with quality-control, supply-chain, and distribution engineering. The public-health system is an engineered surveillance, response, and prevention infrastructure. The pandemic-response system is an engineered logistics-and-manufacturing surge capacity. The chapter takes the healthcare system as a civilization-scale engineering domain and treats the engineering profession's contributions and responsibilities across the layers.

\section{Healthcare as engineered system}\pathtag{core}

A national healthcare system is the integration of: primary-care delivery (clinics; community providers; emergency medical services); secondary-and-tertiary care (hospitals; specialist centres); pharmaceutical supply (manufacturers; distributors; pharmacies); diagnostic infrastructure (laboratories; imaging centres); medical-device manufacturers and service providers; public-health agencies; insurance-and-payment infrastructure; medical-education and workforce-development institutions. The system's spending in developed countries ranges from $\approx 9\%$ of GDP (OECD median) to $\approx 18\%$ of GDP (US extreme); the workforce is typically $5\%$-$15\%$ of the civilian workforce. The engineering profession's contributions are distributed across the system: biomedical engineering for devices; hospital engineering for facilities; industrial engineering for operations; public-health engineering for water and sanitation; informatics and data engineering for electronic-health-record systems.

The system's outcome metrics are population health (life expectancy; healthy life expectancy; disability-adjusted life years; preventable deaths; maternal and infant mortality), patient-experience (access; satisfaction; wait times), and cost. The cross-national variation in outcomes-per-dollar is large: the USA, with the highest per-capita spending, ranks middle-to-low on outcomes-per-dollar; Japan, much of Western Europe, and selected developing-world systems (Costa Rica, Cuba, Sri Lanka) achieve comparable or better outcomes at substantially lower per-capita cost. The engineering interpretation is that healthcare-system performance is principally an integration-and-coordination problem rather than a per-component-technology problem; the systems that perform best are those whose components are well-coordinated, well-funded for the unglamorous primary-care and public-health functions, and free of the perverse incentives that misallocate effort.

\begin{archetype}[Scaling: hospital throughput]
A typical $500$-bed hospital serves a population of $\approx 100{,}000$ in its primary catchment, with $\approx 20{,}000$ inpatient admissions/yr (average length of stay $4$-$5$ days), $\approx 50{,}000$ emergency-department visits/yr, $\approx 200{,}000$ outpatient encounters/yr, $\approx 5{,}000$ surgical procedures/yr, $\approx 500$ births/yr in a hospital with obstetric services. The engineering scope: capacity planning (bed mix, OR mix, ED capacity, ICU capacity); throughput optimisation (flow management, discharge planning, transfer protocols); resilience (surge capacity, infection-control isolation, disaster-response capability); efficiency (resource utilisation; staff scheduling; equipment maintenance). The hospital is one of the most operationally complex engineered facilities in continuous use.
\end{archetype}

\begin{estimation}
\textbf{Question.} Estimate the daily oxygen requirement of a $500$-bed hospital during routine operations and during a pandemic-surge scenario.

\textbf{Reasoning.} Routine baseline: $\approx 100$ patients on supplemental oxygen at any time, at $\approx 4\,\text{L/min} \times 60 \times 24 \approx 6\,\text{m}^3$ of oxygen per patient-day; total $\approx 600\,\text{m}^3$/day. Pandemic surge (e.g., COVID-19 peak): $\approx 200$-$300$ patients on supplemental oxygen with $\approx 30\%$ on high-flow (e.g., $40\,\text{L/min}$), so the total rises to $\approx 3{,}000$-$5{,}000\,\text{m}^3$/day, an $5$-$10\times$ surge. Hospital oxygen plants are sized for $\approx 2$-$3\times$ routine; the surge exceeds plant capacity. The contemporary engineering responses: VPSA (vacuum pressure-swing adsorption) plants for on-site oxygen generation; bulk-liquid supply contracts with surge clauses; portable concentrators for distributed supply. The 2020-2021 COVID-19 oxygen-supply crises in many developing-world hospitals (India May 2021; Brazil 2021) illustrate the failure mode when the surge planning was inadequate.
\end{estimation}

\section{Hospital design and operation}\pathtag{standard}

Hospital design is an established subdiscipline of building engineering with strong codified standards (FGI Guidelines in the US; HBN in the UK; Department of Health technical memoranda; equivalent guidance in other countries). Key engineering requirements: infection control (HVAC pressure relationships between isolation rooms and adjacent spaces; HEPA filtration; ventilation rates of $6$-$15$ air changes per hour depending on space type); medical gases (oxygen, vacuum, medical air, nitrogen, CO$_2$; piped from central plants with redundancy); life-safety power (essential-power generators with $10$-second start-up; uninterruptible power for life-critical equipment); patient-safety design (fall prevention; medication-safety; equipment-and-monitor visibility from nurse stations); fire-safety with smoke compartmentation; security; data and communications.

Operationally, the hospital is a continuous-process facility. Patient-flow management combines admission scheduling, bed-management, surgical-scheduling, discharge planning, and emergency-and-elective balance. The constraint is typically the integration: a hospital can be at full bed capacity while the ED is gridlocked because discharges are stalled by lack of post-acute placement; a hospital can have empty beds and unscheduled ORs because the surgeon-and-anesthesia schedules are not coordinated. The industrial-engineering profession contributes substantially to hospital operations through methods derived from manufacturing operations research (lean process design; theory-of-constraints; queueing models for bed management).

\begin{principle}[Hospital as continuous flow system]
A hospital is a continuous-flow system in which patients are the units flowing through resources (beds, ORs, ED bays, imaging modalities, lab capacity, staff time). The throughput is constrained by the slowest resource and by the synchronisation across resources. Engineering improvements that focus on individual resources (more ED bays) without addressing the upstream and downstream constraints (admission process; discharge process) typically produce smaller throughput improvements than predicted. The systems-engineering approach is to identify the binding constraints, address them in sequence, and measure the throughput response.
\end{principle}

\section{Medical-device infrastructure revisited from Volume VI}\pathtag{standard}

Volume VI Chapter 9 introduced medical-device engineering as a domain. At civilization scale, the engineering question is the deployment, maintenance, and obsolescence of the medical-device stock. A typical $500$-bed hospital operates $\approx 10{,}000$ medical devices; replacement-life ranges from $5$ years (some monitors) to $20+$ years (imaging equipment). The capital intensity is substantial: imaging equipment alone (MRI, CT, ultrasound, nuclear medicine, mammography, fluoroscopy) is typically $\$50$-$\$200$ million per hospital. The maintenance and biomedical-engineering workforce required to keep this stock operating is typically $\approx 50$ FTE per $500$-bed hospital.

The medical-device safety regime, which the Therac-25 case (\cite{paper:leveson-turner1993}) and Volume X Chapter 9 discussed in detail, applies the same control-system, software, and regulatory engineering at civilization scale: FDA 510(k) and PMA pathways in the US; CE marking in the EU under MDR; equivalent regimes elsewhere. The engineering profession's responsibility includes the design, the manufacturing-quality system, the post-market surveillance for unanticipated failure modes, the recall processes, and the engagement with the clinical-workflow context in which the devices actually operate.

\section{Public-health engineering}\pathtag{core}

Public-health engineering is the application of engineering to the population-level determinants of health. The historic contributions: clean water supply (Volume XII Chapter 3); sanitation; food safety (HACCP and the food-safety regulatory regime; Volume XII Chapter 5); vector control (mosquito control for malaria, dengue, Zika); occupational-health engineering (ventilation, noise, dust, ergonomics); environmental-quality monitoring (air, water, soil contaminants). The public-health benefits of these engineering interventions have been enormous; the McKeown thesis (that medical-care advances are responsible for a smaller share of the 20th-century mortality decline than public-health-engineering interventions) is the historical case for the centrality of engineering to population health.

The contemporary public-health engineering agenda includes air-quality engineering (PM2.5 management in cities; indoor air quality in schools, offices, and homes); climate-and-health adaptation (heat-island mitigation; pandemic-resilient built environment); non-communicable-disease environment (urban design that supports physical activity; food-environment regulation); occupational-health for the contemporary workforce (heat-exposure for outdoor workers; air-quality for workers in high-PM environments). The engineering profession's collaboration with the public-health profession is one of the project's themes; the integration is not always smooth.

\begin{archetype}[Scaling: public-health-engineering returns]
Public-health-engineering interventions typically have very high benefit-cost ratios at population scale. The classic clean-water investment returns $\approx 4$-$25$ in avoided healthcare costs and productivity gains per $1$ invested; sanitation comparable; lead-paint and lead-pipe abatement substantially more. The high returns reflect the population-scale benefit (every member of the exposed population is potentially affected) and the multi-decade durability (an installed clean-water system serves decades of population). The engineering choice that delivers public-health benefit at the scale of populations is among the highest-leverage interventions the engineering profession contributes to civilization.
\end{archetype}

\section{Pandemic engineering: surveillance, response, manufacturing}\pathtag{core}

A pandemic-response system is an engineered set of capabilities including: pathogen surveillance (clinical-case reporting; sequence surveillance; wastewater surveillance); diagnostic infrastructure (PCR and antigen-test manufacturing-and-distribution; laboratory capacity for testing and sequencing); pharmaceutical manufacturing surge (vaccines; therapeutics; oxygen; PPE); healthcare-system surge (ICU expansion; staff-augmentation; field-hospital capacity); non-pharmaceutical-intervention infrastructure (testing-and-tracing; isolation-and-quarantine facilities; communication and behaviour-change infrastructure); supply-chain resilience for the entire above. The COVID-19 pandemic stress-tested every component of the pandemic-response system in nearly every country.

The vaccine-manufacturing engineering achievement of 2020-2021 was historic: the mRNA-vaccine platforms (Pfizer-BioNTech; Moderna) moved from laboratory candidates in January 2020 to authorised products in December 2020, a timeline that compressed the typical $5$-$10$ year vaccine-development cycle to under $12$ months. The manufacturing scale-up moved from a few thousand doses in early production to $10$-$15$ billion doses cumulatively administered worldwide by mid-2022. The engineering contribution combined the biological and chemical engineering of the lipid-nanoparticle delivery system, the manufacturing-process engineering of large-scale mRNA synthesis, the cold-chain engineering of $-70\degC$ ultracold distribution, the supply-chain engineering of the pivotal raw materials and equipment, and the global-distribution engineering through COVAX and bilateral channels.

The pandemic-supply-chain failures were also instructive. The 2020 PPE shortage \cite{paper:cohen-pandemic-supply-2020} revealed the global concentration of mask and ventilator manufacturing in a few Asian countries, the just-in-time inventory policies that left strategic stockpiles inadequate, and the export-restriction dynamics that locked countries out of supply when they needed it most. The Indian oxygen crisis of May 2021 revealed the oxygen-supply-chain fragility at population scale. The vaccine-distribution-equity gap revealed the systemic incentives that delivered vaccines to high-income countries faster than to low-income countries despite manufacturing capacity that could have served both. Each failure is an engineering-and-institutional case for the design improvements that subsequent pandemic-preparedness strategies have begun to address.

\section{Healthcare equity engineering-first}\pathtag{standard}

Healthcare equity is the differential access to and quality of healthcare across population groups within and across countries. The within-country equity question in the US is well-documented: African-American maternal mortality is $\approx 3\times$ White maternal mortality (per CDC); life expectancy varies by $\approx 20$ years across US neighbourhoods within a few miles of each other; access to specialist care depends substantially on insurance coverage and geographic location. The cross-country equity question: a child born in Norway has a life expectancy of $\approx 83$ years; a child born in the Central African Republic, $\approx 55$ years. The engineering profession's role in addressing these gaps is not the totality of the answer but is a substantial component: the deployment of mid-level provider-supported clinic models in underserved areas; the design of low-cost diagnostic and treatment devices for resource-limited settings (the Volume VI Chapter 9 case examples include the Embrace warming sleeping bag for neonatal hypothermia in low-resource settings); the supply-chain engineering for vaccine and essential-medicine delivery to remote and low-income populations; the data-engineering infrastructure for equity measurement and reporting.

\section{Failure: healthcare-system failures}\pathtag{core}

The chapter's failure cases each illustrate a distinct mode of healthcare-system failure.

The Therac-25 software-and-control-system failure \cite{paper:leveson-turner1993} discussed in Volume X Chapter 9 is the classic case of medical-device safety failure: software defects that produced overdose conditions; absent interlocks; poor reporting that delayed pattern-recognition. The institutional response was the medical-device-software regulatory regime that subsequently emerged; the engineering lesson is that medical devices with potential for catastrophic harm require software-engineering rigour that the early-1980s field had not yet developed.

The Mid Staffordshire NHS Foundation Trust failure of 2005-2009 \cite{acc:francis-mid-staffs-2013} is the case of organisational-and-cultural failure at hospital scale: chronic understaffing, poor patient care, regulatory failure to detect, target-driven management that prioritised financial metrics over patient outcomes, and a hospital culture that suppressed clinical-staff concerns. The Francis Report (the public inquiry's report, 2013) documented the failure mode and triggered systemic NHS reforms. The engineering and institutional lesson: a healthcare-delivery organisation can fail at the cultural-organisational scale in ways that look qualitatively similar to the Vol X Chapter 10 organisational-failure cases (Challenger; Bhopal; Chernobyl).

The COVID-19 PPE-and-ventilator supply-chain crises \cite{paper:cohen-pandemic-supply-2020} are the case of supply-chain-resilience failure at civilization scale: a global supply chain that was efficient for normal demand and brittle for surge demand; strategic stockpiles that were inadequate or had been allowed to expire; export-restriction dynamics that prevented the global-supply solution from operating; just-in-time inventory policies that had hollowed out the buffer. The engineering response: distributed manufacturing capacity (the Project Defend in the UK; the Strategic National Stockpile reform in the US); the rebuilding of strategic stockpiles; the rebuilding of domestic manufacturing for critical medical supplies; the international agreements (the WHO Pandemic Agreement under negotiation as of 2024) intended to prevent recurrence.

\begin{principle}[Healthcare resilience is engineered, not assumed]
Healthcare-system resilience to high-consequence low-frequency events (pandemics; mass-casualty events; major utility failures; cyber incidents) is the result of engineered investments in surge capacity, redundant supply chains, strategic stockpiles, cross-training, and inter-organisational coordination. None of these investments produces visible benefit in normal operation; all are testable only by the events they are designed against. The engineering profession's contribution: making the resilience investments specific, quantified, and monitorable so that the political-economic case for them can be sustained outside the crisis episodes.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Healthcare system resilience assessment]
Choose a national or regional healthcare system with available data. Document the system's resilience properties: ICU capacity per capita; oxygen-supply system; ventilator stockpile; PPE stockpile; pharmaceutical supply-chain dependencies; clinical-workforce surge capacity; public-health surveillance infrastructure. Assess the resilience against a defined scenario (a pandemic at COVID-19 scale; a major weather disaster; a multi-week utility outage). Identify the specific failure modes that would occur; estimate the time-to-failure; identify the interventions required and their cost. Document the just-resilience strategy that ensures the resilience investments benefit equitably across the population.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A $500$-bed hospital has average length of stay $4.5$ days and average occupancy $85\%$. Compute the annual admissions and the daily admission rate.
\end{exercise}

\begin{exercise}
A hospital's essential-power generator is rated $2{,}000\,\text{kW}$ and must supply the life-safety, critical, and equipment branches at full load for $96$ hours per code. Compute the diesel-fuel storage required at typical generator efficiency.
\end{exercise}

\begin{exercise}
A regional vaccine cold-chain serves $5$ million people at $2$ doses per year at $-20\degC$ to $+8\degC$ for storage and transport. Compute the cold-storage volume required, the daily-throughput capacity at distribution centres, and the energy demand.
\end{exercise}

\begin{exercise}
A pandemic-PPE strategic stockpile aims for $90$ days of surge demand for healthcare-worker N95 masks across a population of $50$ million. Estimate the surge demand (masks per healthcare worker per day during pandemic), the stockpile size, the storage volume, the annual rotation cost.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive an M/M/c queueing model for emergency-department patient flow. Identify the relationship between average wait time, server utilisation, and arrival-rate variability.
\end{exercise}

\begin{exercise}
Derive the ventilation-rate requirement for an airborne-infection isolation room as a function of pathogen-emission rate, target equivalent-quanta concentration, and room volume (Wells-Riley model).
\end{exercise}

\begin{exercise}
Derive the SIR (susceptible-infectious-recovered) epidemic model and identify the basic reproduction number $R_0$ as a function of contact rate, transmissibility, and infectious period.
\end{exercise}

\begin{exercise}
Derive the cost-effectiveness threshold for a healthcare intervention as a function of the willingness-to-pay per QALY and the intervention's incremental cost and effectiveness.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the global pandemic-preparedness investment required to bring all WHO member states to the JEE (Joint External Evaluation) high-capacity tier across the core capacities (surveillance, laboratory, workforce, supply chain).
\end{exercise}

\begin{exercise}
Estimate the global ICU bed deficit in low-income countries relative to a WHO-recommended level. Identify the engineering and workforce investments required.
\end{exercise}

\begin{exercise}
Estimate the per-capita health benefit (in life-years gained or DALYs avoided) of a public-health-engineering intervention in air-quality (PM2.5 reduction by $10\,\mu\text{g/m}^3$ in a major city) over $30$ years.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate a hospital ED during a multi-day surge event. Identify the patient-flow bottlenecks; identify the interventions (additional staff, alternative discharge destinations, satellite care) that would have the largest effect.
\end{exercise}

\begin{exercise}
Simulate a pandemic-response under different preparedness scenarios. Sweep the preparedness-investment parameter and identify the configurations that contain the outbreak versus those that allow it to spread.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design an oxygen-supply system for a $1{,}000$-bed hospital with $3\times$ pandemic-surge capacity. Specify the on-site generation, the bulk-liquid storage, the redundancy, the distribution piping, the monitoring system.
\end{exercise}

\begin{exercise}
Design a regional pandemic-surveillance system combining clinical reporting, laboratory sequencing, and wastewater monitoring. Specify the data-flow, the threshold-trigger logic, the response-coordination interface.
\end{exercise}

\begin{exercise}
Design a low-cost diagnostic-and-treatment package for the leading causes of preventable mortality in a low-income country. Specify the device-and-pharmaceutical inventory, the workforce-and-training, the distribution-and-supply system, the cost-and-coverage outcomes.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: the engineering profession has a substantial contribution to healthcare-system performance that is not currently being made; the integration of engineering into healthcare-system design is the highest-leverage opportunity. Argue the opposite. Identify the empirical evidence.
\end{exercise}

\begin{exercise}
Discuss the COVID-19 pandemic-supply-chain failure. Identify the engineering and institutional contributions to the failure; identify the engineering and institutional interventions that would prevent recurrence.
\end{exercise}

\begin{exercise}
Discuss the Mid Staffordshire NHS failure as an organisational-engineering failure. Identify the parallels and differences with the Vol X Chapter 10 organisational-failure cases; identify the engineering interventions for healthcare-organisation resilience.
\end{exercise}

\begin{exercise}
Argue: healthcare equity is an engineering responsibility because the engineering profession designs the systems whose distribution determines who gets what care. Argue the opposite. Identify the implications for engineering education and professional practice.
\end{exercise}
