\chapter{What design is}
\label{vol11:ch01}

\epigraph{Design is the integration of what physics will permit, what people need, what the engineer knows, and what the engineer commits when the first three are not enough.}{}

\chapmeta{Half-life: medium-life. Archetypes: ethics, optimisation. Exercise target: 15. Project track: hybrid.}

\noindent
Design is what engineering is. The earlier volumes of this book gave the reader the tools, the materials, the mathematics, the failure literacy. This chapter is the discipline that integrates them. Design is not the application of formula to specification; it is the work of producing an artefact that did not exist before and that meets a need that was previously unmet, under constraints that physics and economy and regulation impose, with judgment that the formula does not contain. This chapter develops what design is, what its discipline looks like, what it is not, and what the cases of design-without-need look like.

\section{Design as the synthesis of constraint, want, knowledge, judgment}\pathtag{core}

A design is the resolution of four kinds of input. The first input is \textbf{constraint}: the physics, materials, geometry, energy budget, cost budget, schedule, regulatory requirements, and other limits that the design cannot violate. Constraints are non-negotiable in principle; they may be relaxed by re-scoping the problem, but they cannot be wished away. A design that violates a constraint is not a design; it is a sketch that will not survive contact with reality.

The second input is \textbf{want}: the need that the design is intended to serve. Wants are stated by the user, the client, the community, the future. They are sometimes articulate (an explicit requirement document); they are sometimes inarticulate (a user knows that the current solution is uncomfortable but cannot describe why); they are sometimes contradictory (different users want different and incompatible things); they are sometimes evolving (the want at the start of the design will not be the want at the end). The discipline of requirements engineering (Chapter 2) and user research (Chapter 3) is the discipline of turning want into engineering input.

The third input is \textbf{knowledge}: what the engineer brings to the problem. Domain knowledge (the engineer knows how this class of problem is usually solved); analytical knowledge (the engineer can compute the consequences of design choices); craft knowledge (the engineer has built similar things and remembers what worked); social knowledge (the engineer knows who in the organisation can help; what the regulators expect; what the manufacturer can actually produce). Knowledge is distributed: the engineer's own knowledge plus the engineering team's plus the broader profession's plus the literature's. The design draws on knowledge across all these scales.

The fourth input is \textbf{judgment}: the commitment the engineer makes when the previous three do not determine the answer. Most non-trivial design problems are under-determined: the constraints permit multiple solutions, the wants are compatible with multiple concepts, the knowledge supports multiple analyses. The engineer chooses. The choice is judgment; it is informed by the other three inputs and constrained by them, but it is not derived from them. The engineer who is uncomfortable with judgment is uncomfortable with design; the engineer who pretends judgment can be eliminated by more analysis is mistaken about what design is.

The synthesis of the four is what produces the design. The synthesis is iterative: the engineer proposes a candidate solution, evaluates it against constraints, checks it against wants, refines it with knowledge, and commits the judgments that complete it; the evaluation reveals new constraints or new wants or new knowledge; the engineer iterates. The number of iterations depends on the problem; one is unusual, ten is common, hundreds are not unprecedented for difficult problems.

\begin{principle}[The four inputs of design]
A design is the synthesis of constraint, want, knowledge, and judgment. The discipline of design is the discipline of being honest about each: knowing which constraints are physical and which are negotiable; knowing which wants are documented and which are assumed; knowing which knowledge is the engineer's own and which is borrowed; knowing where judgment was committed and recording it for accountability. A design that pretends judgment was not committed pretends that it was derived rather than designed.
\end{principle}

\section{The design process: stages and iterations}\pathtag{core}

The design process has been formalised in many traditions: the V-model in systems engineering (specification on one side; verification on the other); the waterfall in classical software engineering (sequential stages with formal handoffs); the spiral model (iterative cycles of risk identification, prototype, evaluation); agile in modern software engineering (continuous iteration with frequent delivery); the German VDI 2221 in mechanical design (stage-gated with prescribed deliverables); the IDEO design-thinking framework (empathise, define, ideate, prototype, test). The frameworks differ in detail and in the domains they fit best; they share a substantive structural similarity.

The common stages, abstracted across frameworks:

\textbf{Problem definition.} The engineer characterises what is actually wanted, by whom, under what conditions. The deliverable is a problem statement specific enough to act on. The stage is the precondition for the rest; problems that are not adequately defined produce designs that solve the wrong problem.

\textbf{Requirements specification.} The wants are translated into engineering requirements: measurable, testable, prioritised statements of what the design must do. The deliverable is a requirements document. Chapter 2 develops this stage in depth.

\textbf{Concept generation.} The engineer (or the team) generates multiple candidate solutions to the problem. The discipline is breadth: multiple concepts are explored before any one is committed to. Chapter 4 develops conceptual design.

\textbf{Concept selection.} The candidate concepts are evaluated against the requirements; the best one (or a small set of finalists) is selected for detailed development. The selection is not always the highest-scoring concept; the judgment of the design team weighs factors that the scoring matrix does not capture.

\textbf{Detailed design.} The selected concept is developed in detail: dimensions, materials, tolerances, interfaces, components specified to the level that manufacturing can act on. Chapter 5 develops detailed design.

\textbf{Analysis and simulation.} The detailed design is analysed against the requirements: structural analysis, thermal analysis, control analysis, lifecycle analysis, cost analysis, manufacturability analysis. Chapter 6 covers simulation.

\textbf{Prototyping.} A physical or simulated instance of the design is built and operated to test assumptions, validate analyses, and surface issues that the analyses missed. Chapter 7 covers prototyping.

\textbf{Testing and validation.} The prototype (or the production design) is tested against the requirements: does it meet what was specified? Chapter 8 covers testing.

\textbf{Review and approval.} The design is reviewed by qualified persons (peers; senior engineers; safety engineers; regulators) and approved for the next stage. Chapter 9 covers reviews.

\textbf{Release.} The approved design is released to manufacturing, construction, or deployment. The release marks the transition from design to realisation; subsequent changes are governed by change-control rather than redesign.

The stages are conceptually sequential but operationally iterative. Issues surfaced in testing send the design back to detailed design or to concept selection; analysis results may require re-specification; user feedback may revise requirements. The iteration is not failure of the process; it is the process. The engineer who treats iteration as exceptional and one-pass-through as normal has misread the discipline.

The cost of changes rises sharply with stage: a change at requirements is cheap; a change at concept selection is moderate; a change at detailed design is expensive; a change after release is very expensive (recalls, rework, contractual penalties, reputational cost). The principle of "shift left" (catch issues as early in the process as possible) drives the investment in requirements work, user research, conceptual exploration, and early prototyping; the alternative (catch issues late) is the dominant historical pattern and is reliably more expensive in the aggregate.

\begin{archetype}[optimisation]
Design is the optimisation archetype's working expression. The engineer is searching the design space for solutions that satisfy the constraints, meet the wants, and use the knowledge well. The optimisation is multi-objective (multiple wants, sometimes incompatible), constrained (physics, cost, schedule, regulation), high-dimensional (many design variables), and contested (different stakeholders weight the objectives differently). The discipline of design is the discipline of structured optimisation under those conditions.
\end{archetype}

\section{The design space and how to navigate it}\pathtag{standard}

The design space is the set of all possible designs that satisfy the constraints. For non-trivial problems, the space is large; for many problems, it is too large to enumerate. The engineer's task is not to enumerate the space but to navigate it: to identify promising regions, to characterise the tradeoffs across them, to select a candidate that warrants detailed work, and to refine the candidate to the deliverable.

The navigation has several characteristic moves:

\textbf{Decomposition.} The problem is split into subproblems whose design spaces are smaller and whose interfaces are explicit. A bridge design is decomposed into deck design, span design, foundation design, expansion-joint design, drainage design, lighting design; each subproblem is approached separately, with the interfaces (loads transmitted; spatial accommodations; budget allocations) explicit. Decomposition is the discipline by which large design problems become tractable.

\textbf{Reuse.} An existing design (or design pattern) is adapted to the current problem. The reuse may be at the level of a complete product (a previous bridge of similar span is studied as a precedent), a subsystem (a previously-developed control loop is incorporated), a component (an off-the-shelf bearing is selected), or a pattern (the cantilever construction method is adapted). Reuse compresses the design effort by drawing on prior work; it is the dominant move in working engineering practice.

\textbf{Analogy.} A solution from a different domain is adapted to the current problem. The aerospace honeycomb structure adapted to floor panels in trains; the automotive suspension geometry adapted to the prosthetic limb; the database transaction model adapted to the cash-machine system. Analogies open the design space beyond the engineer's specific domain; they require judgment about which features of the analogue transfer and which do not.

\textbf{Bounding.} The engineer establishes upper and lower bounds on the achievable: the lower bound by demonstrating that a simple solution works; the upper bound by computing what the physics permits. The space between the bounds is where the design lives; characterising the bounds before searching the space saves wasted effort.

\textbf{Pareto exploration.} For multi-objective problems, the engineer identifies the Pareto frontier (the set of designs where improving any one objective requires sacrificing another) and selects from along it. The selection point on the frontier is judgment; the frontier itself is engineering analysis. The discipline distinguishes "the design is optimal" (it is on the Pareto frontier) from "the design is the right point on the frontier" (the second claim is the judgment one).

\textbf{Constraint elimination.} A binding constraint is examined for whether it is actually binding: can the materials budget be increased; can the regulatory requirement be challenged; can the user accept a different solution to the underlying need. Constraint elimination opens the design space; it is the move that produces the breakthrough designs that ordinary search would miss.

The navigation is a discipline; it is taught through case studies, through working with experienced designers, through repeated practice. The engineer who can navigate the design space competently is the engineer whose designs reliably approach the achievable; the engineer who cannot is the engineer whose designs reliably fall short of it.

\section{Design as conversation with reality}\pathtag{standard}

The previous sections present design as a structured activity with stages and methods. The activity has an underlying epistemic structure that deserves explicit statement: design is a conversation with reality. The designer proposes; reality responds; the designer revises; reality responds again; the conversation continues until the design is committed.

Reality's responses come through multiple channels. The analysis (structural, thermal, control, cost, lifecycle) responds with predictions about whether the design will work. The simulation responds with more detailed predictions, often visualised. The prototype responds with behaviour the designer can observe. The test responds with measurements against the requirements. The user responds with feedback about whether the design serves the want. The manufacturer responds with information about whether the design can be built. The regulator responds with judgements about whether the design satisfies the regulatory framework. The market responds with adoption or rejection.

Each channel has different latency, fidelity, and cost. Analysis is fast and cheap but assumes a model of the design that may differ from the actual artefact. Simulation is slower and more expensive but uses richer models. Prototyping is slower and more expensive still but produces something close to reality. Testing on the production design is the most expensive and most authoritative. User feedback can be obtained at all stages with varying fidelity (early concept sketches; mid-stage prototypes; late-stage products). The designer chooses the channels that produce the most informative responses at the lowest cost; the choice is itself a design decision.

The conversation requires that the designer listen. The defensive posture (the design is correct; reality is wrong; the analysis must be re-run until it agrees with the design) produces designs that fail in service. The receptive posture (the design is a hypothesis; reality is the evidence; the design must change when the evidence requires) produces designs that converge to working artefacts. The cultural distinction is substantial; the conversation-with-reality framing is the explicit articulation of the receptive posture.

The conversation has limits. Reality does not always respond in time for the response to influence the design; some designs must be committed under residual uncertainty about how they will perform. The discipline is then the discipline of designing for uncertainty: design margins; staged commitment; reversibility; instrumentation that will reveal the as-built behaviour after release. The conversation continues across the design's lifecycle; the in-service feedback feeds into the next generation of design.

\begin{principle}[Design is a conversation]
Design is iterative engagement with reality through analysis, simulation, prototype, test, user feedback, and in-service evidence. The designer who treats the design as a commitment to be defended produces designs that fail; the designer who treats the design as a hypothesis to be tested produces designs that work. The discipline of design is the discipline of staying in the conversation through commitment.
\end{principle}

\section{What design is not: gilded analysis, decoration, marketing}\pathtag{standard}

The boundary of design as engineering discipline runs along several edges. Activities that overlap with design but are not design:

\textbf{Gilded analysis.} The engineer performs extensive analysis on a design that was already committed; the analysis confirms what the design assumed; no design choices are altered by the analysis. The activity is paper engineering: it produces documents that look like design work but contributes nothing to the artefact's eventual performance. The discipline of design distinguishes analysis that informs design decisions from analysis that justifies decisions already made.

\textbf{Decoration.} The engineer (or, more commonly, the industrial designer; or, more commonly still, the marketing department) adds features to a design that do not serve a need: badges, ornament, complexity that signals capability without enabling it. Decoration is the appearance of design without the substance. The discipline of design distinguishes features that serve user need from features that serve only the engineer's preference or the marketing department's narrative.

\textbf{Marketing.} The activity of communicating what a design does to potential users and buyers is marketing; it is a legitimate activity but it is not design. The conflation arises when marketing requirements (the design must support the marketing narrative; the design must look like it does X even though it does Y) override engineering judgment. The discipline of design protects the engineering judgment from marketing pressure; the appropriate channel for marketing requirements is the requirements specification (Chapter 2), where they enter as one input among others.

\textbf{Specification compliance.} A design that meets every specification but does not serve the underlying need is a failure of design that the specification did not catch. The discipline of design includes the discipline of checking whether the specification captured the need adequately; the specification is the engineering encoding of the need but it is not the need itself.

\textbf{Optimisation of fixed concept.} The engineer takes a committed concept and optimises its parameters within the concept's structure; the activity is parameter optimisation, not design. Parameter optimisation is necessary but is not the dominant design work; the more impactful design moves are at the concept level (Chapter 4) where the parameter space is being established.

\textbf{Pattern application.} The engineer applies a well-known design pattern to a new problem; the application is reuse rather than design. Pattern application is legitimate and often the right move; the design discipline asks whether the pattern fits the problem and whether modifications are required.

The boundary is not always sharp. Some activities are partly design and partly something else; the discipline is the discipline of being honest about which is which. An engineer who is doing parameter optimisation should call it parameter optimisation rather than design; an engineer who is justifying a committed design through analysis should call it justification rather than design. The honesty preserves the value of the design discipline by reserving the term for the substantive activity.

\section{Failure: design that did not start from need}\pathtag{core}

The most consequential design failures are designs whose starting point was not a need but something else: a technology in search of an application; a feature set that the development organisation wanted to ship; a market position the company wanted to defend; a procurement requirement that no end-user requested. The designs may be competent in execution; their failure is at the starting point. The artefact, however well-engineered, does not serve any actual user.

\textbf{Technology-push failures.} A new technology becomes available; a design organisation searches for problems the technology can be applied to. The search sometimes finds genuine needs that the technology addresses well; it sometimes finds shallow applications that do not actually need the technology and that would be better served by simpler solutions. The early commercial-aircraft electronic flight bag deployments of the 2000s were partly technology-push: the available tablet hardware was applied to pilot information needs, with some genuine improvements over paper and some applications that paper served at least as well. The post-implementation lessons drove the more thoughtful subsequent designs.

\textbf{Feature-push failures.} A product organisation has a list of features it wants in the next release; the design work is to incorporate the features. The features may not have been requested by users; they may have been generated by competitive analysis ("the competitor's product has feature X; we must too"), by the engineering team's interests, or by management's intuition. The resulting designs are complex, expensive to maintain, and produce inferior user experience compared to simpler designs that address actual needs. The software industry's "feature creep" pattern is the canonical case.

\textbf{Procurement-driven failures.} A government or corporate procurement specifies requirements that no operational user requested; the design is built to the procurement specification; the operational user receives a product that does not serve their need. The military procurement record contains many examples; the civilian large-IT procurement record contains many more. The remediation typically requires expensive re-specification after the initial design has been delivered.

\textbf{Vanity-driven failures.} A design organisation produces an artefact that demonstrates the organisation's capability without serving an external need: an architectural building that is unusable as a building but photographs well; a vehicle that wins design awards but does not transport effectively; a software product that demonstrates technical sophistication but is not used. The artefact serves the organisation's vanity; the user need it putatively addresses is not actually addressed.

\textbf{Specification-trap failures.} A design is built to a specification that captured the need at a point in time; the need has evolved; the specification has not. The design serves the original specification competently but does not serve the current need. The remediation is constant re-engagement with the user to verify that the specification continues to capture the need; the discipline is the user-research discipline of Chapter 3.

The diagnostic move that catches these failures is the question "what need is this design serving?" asked with discipline at every design review. If the answer is not crisp, specific, and verifiable, the design has a starting-point problem that no amount of execution will overcome. The discipline of design begins with the discipline of need; the chapters that follow develop the disciplines that extend from there.

\begin{principle}[Design starts from need]
A design that does not start from a verified need is a design that may execute well but cannot succeed. The first discipline of design is the discipline of articulating the need clearly enough that the design can be judged against it. Designs that start from technology, feature, procurement, or vanity may produce artefacts; the artefacts do not constitute design in the sense this volume develops.
\end{principle}

\begin{project}[Ten objects, ten needs]
List ten engineered objects in the reader's environment: tools, appliances, vehicles, buildings, software systems, infrastructure components. For each object, write one paragraph identifying: the original design need (what was this object designed to do?); the actual current use (what is this object actually used for?); the divergence (where do need and use differ?); the design opportunity (what could a redesign address?). Deliverable: a ten-page document with one page per object, plus a one-page synthesis identifying patterns across the ten.

\textbf{Hazard.} The exercise often reveals that the reader does not know the original design need for objects they use daily; the discipline of inquiry is part of the project. Where the original need is not recoverable from public sources, the reader should note the gap rather than speculate.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
For a four-element optimisation under constraint (cost $\leq C$; mass $\leq M$; reliability $\geq R$; manufacturability score $\geq Q$), compute the number of constraint boundaries in the four-dimensional design space. Identify the number of facets, edges, and vertices the constraint polytope has under the convex-bounded approximation.
\end{exercise}

\begin{exercise}
A design problem has 6 candidate concepts to evaluate against 8 requirements. Construct a Pugh decision matrix (concepts as columns; requirements as rows; entries as $+$, $0$, or $-$ relative to a datum concept). Compute the total scores for an example scoring assignment and identify the sensitivity of the ranking to weighting choices.
\end{exercise}

\begin{exercise}
The cost of design changes rises with stage of development. If the change cost is $C_0$ at requirements stage, $5 C_0$ at detailed design, $50 C_0$ at prototype, $500 C_0$ at production, and $5000 C_0$ after deployment, compute the cost of a single requirements-stage change vs. an equivalent change discovered after deployment. Discuss the design-process investment implications.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the necessary conditions for a design to be Pareto-optimal in a multi-objective optimisation: the gradients of the objectives must satisfy a specific linear combination at the Pareto-optimal point. Identify how the conditions degenerate when objectives are correlated.
\end{exercise}

\begin{exercise}
For a decomposition of a design problem into $N$ subproblems with interface costs $c_{ij}$ between subproblems $i$ and $j$, derive the conditions under which the decomposition reduces total design effort below the integrated approach. Identify when decomposition is counterproductive.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the number of design iterations a typical mid-complexity engineering project (e.g., a new household appliance) undergoes from requirements to production release. Identify the cost dominance of late-stage iterations.
\end{exercise}

\begin{exercise}
Estimate the fraction of design effort, in a typical product-development organisation, allocated to: requirements; concept; detailed design; analysis; prototyping; testing; reviews; design-for-manufacture. Identify the disparities between organisations that produce successful products and those that do not.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Choose one engineered object the reader uses regularly and that the reader believes is poorly designed. Identify the design constraints, the user wants, the engineering knowledge, and the judgments that produced the current design. Propose a redesign that addresses one specific deficiency. Identify what the redesign would cost in design effort and lifecycle implications.
\end{exercise}

\begin{exercise}
Design a problem-definition checklist that an engineering team can apply at the start of a new project. Identify the questions whose answers must be specific before requirements work begins, and the questions whose answers identify when the team should pause before proceeding.
\end{exercise}

\begin{exercise}
Design a "what need is this design serving?" review-board protocol for design reviews at every stage (concept, detailed design, prototype, validation, release). Identify the specific questions reviewers should ask and the conditions under which the review should block the design from proceeding.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: the engineer who treats design as parameter optimisation of a committed concept has misread the discipline. Identify the conditions under which parameter optimisation is the dominant move (legitimate) versus the conditions under which it substitutes for substantive concept-level design work (illegitimate).
\end{exercise}

\begin{exercise}
Discuss the role of judgment in design. Argue: as analysis tools mature (simulation, AI-assisted design, optimisation algorithms), the scope of judgment shrinks. Argue the opposite: as analysis tools mature, the scope of judgment shifts but does not shrink. Identify the empirical evidence that bears on the dispute.
\end{exercise}

\begin{exercise}
A design organisation operates on technology-push: it identifies new technologies and searches for applications. Discuss the conditions under which technology-push design produces genuine value (the technology addresses needs that simpler approaches do not address) versus the conditions under which it produces shallow applications (the technology is applied to needs that simpler approaches address as well or better).
\end{exercise}

\begin{exercise}
Argue: the dominant cause of design failures in modern engineering is starting-point failures (the design did not begin from a verified need) rather than execution failures (the design began from a need but was executed poorly). Identify the empirical evidence that bears on the position and discuss its implications for engineering education and practice.
\end{exercise}

\begin{exercise}
Take one engineering project from the reader's recent experience or from a public case study. Identify the four design inputs (constraint, want, knowledge, judgment) as they appeared in the project: where each input came from; how the inputs were balanced; where the judgment was committed and recorded. Identify one input that was under-attended to in the actual project and what a stronger attention to that input would have changed.
\end{exercise}
