\chapter{Cascading failure in networks}
\label{vol10:ch07}

\epigraph{The networks that bind us also propagate our worst events.}{}

\chapmeta{Half-life: durable. Archetypes: scaling, failure, transport, coupling. Exercise target: 25. Project track: simulation.}

\noindent
Cascading failure is the propagation of failure through a coupled network: an initial event stresses adjacent components; some fail; their failures stress others; the cascade expands. The phenomenon is the dominant systemic-risk mode in modern interconnected systems: power grids, financial markets, internet routing, supply chains, transportation networks. Volume IX Chapter 13 introduced cascading failure in network systems; this chapter expands the treatment as a failure mode with its own dynamics, modelling, and engineering response. The cases include the 2003 North America blackout, the 2008 financial crisis, the 2021 Suez blockage, and the BGP routing leaks of various years.

\section{Why networks cascade}\pathtag{core}

A network without cascading is just a collection of components. Cascading depends on the network's coupling: a component's failure must propagate through the network to other components. The propagation can be physical (load redistribution in a power grid), informational (BGP route changes propagating), economic (price changes propagating through markets), logistical (delays propagating through supply chains), or behavioural (panic propagating through populations).

The conditions for cascading:

\textbf{Coupling:} components affect each other. Without coupling, failures are isolated.

\textbf{Stress proximity to capacity:} the surviving components must approach their capacity limits after the initial failure. Components with comfortable margin absorb the stress without further failure.

\textbf{Positive feedback:} the failure produces conditions that promote further failure rather than damping the event. Negative-feedback systems absorb perturbations; positive-feedback systems amplify them.

\textbf{Insufficient damping:} the timescale of the failure (cascade speed) exceeds the timescale of the response (protective action, operator intervention, market correction). Slow cascades can be interrupted; fast cascades complete before response is possible.

\textbf{Path multiplicity:} there are multiple paths along which failure can propagate; isolating one path does not stop the cascade.

\textbf{Common-mode triggers:} the same external event affects many components simultaneously; the cascade starts with multiple seeds rather than one.

The engineering response is the inverse of these conditions: introduce decoupling between subsystems, operate with margin, design negative-feedback responses, ensure that protective actions are faster than failure propagation, limit path multiplicity through isolation, and address common-mode triggers explicitly.

\begin{archetype}[failure: as cascade through coupling]
A cascading failure is a failure that crosses the coupling between components. The single-component failure is not the failure of the system; the failure of the cascade-spreading mechanism is. The engineering attention shifts from making each component reliable to making the coupling forgiving: islanding capability, capacity margin, isolation between subsystems, fast protective response, diverse paths. The cascade is the structural risk of any tightly-coupled network; the engineering response is to design the coupling, not just the components.
\end{archetype}

\section{Power-grid cascades}\pathtag{standard}

The power grid is the prototypical cascading-failure system: tightly-coupled by physics, operating near capacity for economic reasons, with protective relays designed for local rather than system-wide protection.

The cascade mechanism: a transmission line trips (vegetation contact, equipment failure, lightning); its current is redistributed to adjacent lines; if any of those exceed their thermal capacity, they trip; the cascade continues. The 2003 Northeast blackout illustrated the mechanism: a vegetation contact in Ohio led to relay miscoordination, then to a series of line trips, then to islanding of large grid regions, then to load-loss in the islands.

The protection challenges:

\textbf{Local vs system protection:} local relays trip the affected component; system-wide protection requires inter-utility coordination. The coordination is technically and politically difficult.

\textbf{Reverse-power flow:} when generation is lost, power flows reverse direction; relays designed for one-way flow may mis-coordinate.

\textbf{Voltage collapse:} reactive-power balance degrades; voltages drop; equipment trips on undervoltage; the collapse cascades.

\textbf{Frequency collapse:} active-power balance degrades; frequency drops; generators trip on underfrequency; the collapse continues.

\textbf{Islanding:} when the cascade isolates regions, each island must balance its own generation and load; uncontrolled islanding produces local frequency and voltage excursions.

Engineering responses (developed substantially after the 2003 blackout):

\textbf{Reliability standards:} NERC (North American Electric Reliability Corporation) standards require utility actions for security; compliance is monitored and enforced.

\textbf{Real-time monitoring:} phasor measurement units (PMUs) provide synchronised observations of grid state at high rate; the data supports situational awareness.

\textbf{Under-frequency load shedding:} automatic load shedding when frequency drops; sheds load to preserve generators.

\textbf{Special protection schemes:} pre-planned protective actions for specific contingencies.

\textbf{Operator training:} simulators for cascade-event training; operator awareness of system-wide conditions.

\textbf{Inter-utility coordination:} reliability coordinators with cross-utility authority; mutual aid agreements; coordinated planning.

The 2003 blackout (the canonical case) cost $\sim\$6\text{B}$ and affected $55$ million customers across the US Northeast and Ontario for hours to days; the investigation produced extensive recommendations on standards, coordination, and protection. The 2012 India blackout affected $\sim 620$ million people; the mechanism was similar; the response was similarly extensive.

\section{Financial-system cascades engineering-first}\pathtag{mastery}

Financial systems exhibit cascading failure in a different mode: economic and behavioural coupling rather than physical. The 2008 financial crisis is the canonical large-scale cascade in modern memory.

The mechanism: leveraged institutions hold each other's debt; counterparty risk couples them; the failure of one institution stresses its counterparties; their failures stress theirs; the cascade can take down major institutions and disrupt the broader economy. The 2008 crisis added: mortgage-backed securities of dubious quality; credit-default swaps that concentrated rather than diluted risk; ratings agencies that mislabelled risk; regulatory capture that allowed dangerous practices.

Engineering aspects:

\textbf{Margin requirements:} like power-grid capacity margin, financial institutions hold capital and liquidity reserves. Insufficient margin allows perturbations to cascade; excessive margin reduces system capacity.

\textbf{Counterparty exposure:} like grid coupling, counterparty positions create the cascade pathway. Limits on individual counterparty exposure reduce cascade risk.

\textbf{Central clearing:} replacing bilateral counterparty relationships with central counterparties (CCP) reduces direct exposure but concentrates risk at the CCP. The CCP's failure becomes a systemic risk.

\textbf{Stress testing:} large financial institutions undergo stress tests by regulators; the tests verify the institution's resilience to specific adverse scenarios.

\textbf{Macroprudential regulation:} systemic-risk regulation oversees the system as a whole, not just individual institutions. The Basel Committee on Banking Supervision, the Financial Stability Board, and similar bodies coordinate internationally.

The engineering metaphor is imperfect: financial systems involve human behaviour (panic, herding, irrational expectations) that physical systems do not. But the cascade structure is similar: tight coupling, positive feedback, insufficient margin, common-mode triggers.

\section{Internet routing cascades}\pathtag{standard}

The internet's routing system (BGP) experiences cascading events when route changes propagate widely. The propagation is the system's intended behaviour; the failures occur when the route changes are incorrect or maliciously misleading.

The cascade mechanism: an autonomous system (AS) announces a route to a prefix; neighbouring AS accept the route; they propagate it to their neighbours; the route spreads through the internet. If the announcement is incorrect or attacks legitimate prefix ownership, traffic is misdirected at scale.

Notable cases:

\textbf{2008 Pakistan-YouTube incident:} a Pakistani ISP announced a more-specific YouTube prefix internally to block YouTube domestically; the announcement leaked externally; the route propagated globally; YouTube traffic was directed to Pakistan and dropped; YouTube was offline for hours globally until the leak was withdrawn.

\textbf{2010 China telecom incident:} approximately $15\%$ of internet traffic briefly routed through China Telecom; the cause was a misconfiguration of BGP routes; the impact was hours of mis-routing affecting many destinations.

\textbf{2017 Google route leak via Verizon:} a small ISP's BGP announcements were propagated by Verizon to many networks; Google traffic was disrupted for over an hour. The cause was a customer-side configuration error combined with insufficient route-filtering by Verizon.

\textbf{2018 MyEtherWallet via Amazon Route 53:} attackers hijacked the BGP routes to Amazon's Route 53 DNS service, allowing them to redirect DNS lookups for MyEtherWallet; users were directed to a fake site and lost cryptocurrency. The attack involved BGP-level routing manipulation combined with the lack of DNSSEC for the target domain.

Mitigation:

\textbf{RPKI (Resource Public Key Infrastructure):} cryptographic certificates that authenticate IP-prefix ownership; ASs can validate route announcements against the certificates. Deployment is increasing but not universal.

\textbf{BGPsec:} extends RPKI to path validation (not just origin validation); not yet widely deployed.

\textbf{Route filtering:} ASs filter incoming announcements based on policy; well-known prefixes are protected; suspicious announcements are rejected.

\textbf{Multi-path routing:} avoiding single points of dependency for critical services; using multiple providers and multiple paths.

\textbf{DNSSEC:} cryptographic authentication of DNS responses; protects against DNS-level hijacks (which often combine with BGP hijacks).

The BGP cascade problem is structural: the protocol is policy-flexible (each AS expresses its own routing policy) and that flexibility allows the leaks. Replacing BGP would be enormously disruptive; the engineering response is incremental hardening rather than replacement.

\section{Supply-chain cascades}\pathtag{standard}

Supply-chain cascades occur when disruption in one supplier propagates through the dependent supply chain. The 2011 Tohoku earthquake and tsunami in Japan disrupted multiple supply chains: automotive (Japanese suppliers of specialised components), electronics (Japanese suppliers of semiconductors and other components), pharmaceuticals (Japanese suppliers of intermediates). Industries with sole-source dependencies experienced months of disruption.

The 2021 semiconductor shortage was a multi-cause cascade: pandemic-related demand-pattern shifts (automotive demand collapse then recovery; consumer electronics demand surge); fab capacity tightness from prior under-investment; weather events (Texas freeze in February disrupted Samsung fabs); fire (Renesas fab in March); shipping disruptions; geopolitical tensions. The combination produced automotive production halts, electronics product delays, and consumer-product shortages for over a year.

The 2021 Suez Canal blockage by the Ever Given (March 23-29) blocked one of the world's most important shipping lanes for six days. The immediate cause was a single ship grounding due to high winds; the cascade affected approximately $\$10\text{B}$ of trade per day; surrounding supply chains experienced weeks of disruption as the backlog cleared; insurance and operational disputes continued for months.

Engineering principles for supply-chain cascade resilience:

\textbf{Geographic diversification:} suppliers in multiple regions reduce single-region disruption risk.

\textbf{Supplier diversification:} multiple suppliers for critical components reduce sole-source disruption risk.

\textbf{Strategic stockpile:} inventory of critical components provides time to respond to disruptions.

\textbf{Visibility:} tracking the supply chain's status, including second-tier and third-tier suppliers, provides early warning of disruptions.

\textbf{Modular designs:} components designed for multiple-supplier interchangeability reduce single-supplier dependence.

\textbf{Onshoring/reshoring:} bringing production closer to the consumption point reduces transport-disruption risk at the cost of higher labour cost.

The engineering trade-off is the same as elsewhere: resilience costs money; the discipline is matching the resilience investment to the consequence of failure.

\section{Models of cascading failure}\pathtag{standard}

Cascading failures are studied by several modelling traditions:

\textbf{Load-redistribution models:} when a component fails, its load is redistributed to others; the redistribution depends on the network topology and the load-flow physics. Models include the Bak-Tang-Wiesenfeld sandpile model (self-organised criticality), the Motter-Lai model (power-grid cascade), and others. The models predict cascade size distributions and the conditions for large cascades.

\textbf{Percolation models:} the network is gradually removed (nodes or links); at a critical fraction removed, the network's giant component disintegrates. The transition is sharp; below threshold the network remains largely connected, above it is fragmented. Applies to network resilience analyses.

\textbf{Epidemic models:} the failure propagates from node to node like an infection. SIR-style models with parameters (infection rate, recovery rate, network structure) predict outbreak sizes. Applies to BGP leak propagation, software-bug propagation in distributed systems, viral content spread.

\textbf{Threshold models:} a node fails when more than a threshold fraction of its neighbours have failed. The threshold determines the cascade dynamics. Applies to social-influence cascades, financial-contagion models.

\textbf{Agent-based simulations:} explicit simulation of each agent's behaviour and interaction; can capture heterogeneity and complex behaviours that aggregate models miss.

\textbf{Game-theoretic models:} agents respond strategically; defenders may have time to react. Captures the adversarial cases (cyberattacks, market manipulation).

The choice of model depends on the question: load-redistribution for grid cascades, percolation for connectivity, epidemic for propagation, threshold for behavioural cascades, agent-based for complex heterogeneous systems.

\section{Case: 2003 North America blackout, 2008 financial crisis, 2021 Ever Given Suez blockage}\pathtag{core}

Three cascading-failure case studies:

\textbf{2003 North America blackout (August 14):} cascaded from vegetation contact with a transmission line in Ohio through a sequence of line trips, generator islands, voltage and frequency excursions, and load loss; affected $55$ million customers for hours to days; cost $\sim\$6$ billion. The investigation (jointly conducted by US-Canada Power System Outage Task Force) identified the proximate trigger (Ohio vegetation contact), the failure to detect and respond (FirstEnergy's alarm system was non-functional, masking the developing situation), the failure of inter-utility coordination (Midwest ISO operators did not have situational awareness of the cascading event), and the protective relay miscoordination that propagated the cascade. The recommendations led to NERC reliability standards mandatory enforcement, real-time monitoring (PMUs), tree-trimming requirements, and operator training enhancements.

\textbf{2008 financial crisis (September 2008 acute phase):} multiple causes including subprime-mortgage exposure, leveraged housing markets, mortgage-backed-security mispricing, credit-default-swap concentration, ratings-agency failures, regulatory gaps. The acute cascade involved Lehman Brothers' bankruptcy (Sept 15), AIG's near-collapse (Sept 16), money-market-fund "breaking the buck" (Sept 16), credit markets seizing up, government interventions including TARP. The total cost (recession losses, government interventions, productivity declines) ran into trillions. The engineering response included Dodd-Frank Act, Basel III capital requirements, central clearing of derivatives, stress testing, and macroprudential supervision.

\textbf{2021 Ever Given Suez blockage (March 23-29):} the $400$-metre container ship Ever Given ran aground in the Suez Canal, blocking all traffic in both directions for six days. The immediate cause involved high winds and operational decisions; the broader cascade affected approximately $\$10\text{B}$ of trade per day; ships rerouted around the Cape of Good Hope (adding two weeks to voyages); the canal's blockage backlog took weeks to clear; insurance, salvage, and liability disputes continued for over a year. The case illustrates how a single physical event in a chokepoint can cascade globally; the engineering response includes calls for chokepoint diversification (Northern Sea Route, expanded Panama Canal), larger chokepoint capacity, and improved ship handling.

\textbf{Common patterns:}

\textbf{Initial event is local:} the cascade starts from a localised event (line trip, bank failure, ship grounding).

\textbf{Cascade exceeds the local cause:} the propagation goes far beyond the initial event's direct impact.

\textbf{Coupling pathways are non-obvious:} the cascade follows pathways that were not fully anticipated before the event.

\textbf{Recovery takes much longer than the initial event:} the backlog and disruption persist long after the initial cause is cleared.

\textbf{Regulatory response follows:} cascades drive regulatory and standards changes that take years to implement.

\begin{principle}[The decoupling-and-margin principle]
Cascading failure is the structural failure mode of tightly-coupled networks. The engineering response is decoupling and margin: design subsystems with explicit boundaries that limit propagation; operate with capacity margin so that perturbations are absorbed; ensure that protective responses are faster than failure propagation; identify the chokepoints and either harden them or provide alternatives; learn from each cascade and update the design. The networks the engineer builds will cascade if conditions align; the engineering work is making the cascades less frequent, smaller, and recoverable.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Simulate a cascading failure on a small network]
\textbf{Objective.} Simulate a cascading failure on a small network; identify the threshold above which cascades become catastrophic; identify the engineering interventions that increase resilience.

\textbf{Method.} Choose a network: a small power grid (the IEEE 14-bus or 39-bus test case), a small social network, a small supply chain, a small communication network. Implement a load-redistribution or propagation model appropriate to the network type. Inject failures of various sizes; observe the cascade size distribution. Identify the threshold loading or coupling above which cascades become catastrophic. Experiment with engineering interventions: increased capacity, isolation between subsystems, diverse paths, faster response; quantify the improvement.

\textbf{Deliverables.} The network description; the simulation code; the cascade-size distribution; the threshold identification; the intervention experiments and results; a written analysis of the engineering lessons.

\textbf{Hazard.} Software only; the educational value is in the cascade physics.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A power grid has $100$ transmission lines, average load $80\%$ of capacity; a single trip redistributes load equally to $4$ neighbours, increasing their load by $5\%$ each. State whether further trips occur if neighbours' capacities are $90\%$ of original line.
\end{exercise}

\begin{exercise}
A percolation model on a $1000$-node random network has $50\%$ of nodes removed. Compute the expected fraction of the network remaining in the giant component (use the standard random-graph result).
\end{exercise}

\begin{exercise}
A supply chain has $4$ tiers; each tier has $90\%$ survival probability under disruption. Compute the probability of complete supply-chain survival.
\end{exercise}

\begin{exercise}
A BGP route propagation reaches $100\%$ of internet ASs in $\sim 10$ minutes. Compute the average propagation hops per minute given $\sim 70{,}000$ ASs and typical $4$-$5$ hop diameter.
\end{exercise}

\begin{exercise}
A financial system has $20$ major institutions; counterparty exposure is concentrated in $3$ central institutions. State the contagion exposure and the role of central clearing.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the critical loading for a simple load-redistribution model with uniform capacity and uniform redistribution.
\end{exercise}

\begin{exercise}
Derive the percolation threshold for an Erd\H{o}s-R\'enyi random graph.
\end{exercise}

\begin{exercise}
Derive the SIR model's basic reproduction number and the threshold for outbreak vs decay.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the expected cascade size for a power-grid loading of $90\%$ of capacity; assume realistic line-impedance distribution.
\end{exercise}

\begin{exercise}
Estimate the diversification benefit of sourcing $30\%$ of a critical component from a second supplier.
\end{exercise}

\begin{exercise}
Estimate the chokepoint resilience improvement from adding a parallel route with $50\%$ of the original capacity.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate a Motter-Lai cascade on an IEEE 14-bus or 39-bus test grid; identify cascade sizes vs initial loading.
\end{exercise}

\begin{exercise}
Simulate a sandpile (Bak-Tang-Wiesenfeld) model; observe the cascade-size distribution's power-law character.
\end{exercise}

\begin{exercise}
Simulate a financial-contagion model with counterparty exposure; observe the cascade behaviour above and below the threshold leverage.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a chokepoint analysis for a critical supply chain: identify single points of failure, the cost of alternatives, the recommended investments.
\end{exercise}

\begin{exercise}
Design a power-grid cascade-mitigation strategy: monitoring, protective relays, special protection schemes, operator interface.
\end{exercise}

\begin{exercise}
Design a BGP-route-filtering strategy for an ISP that wants to reduce its exposure to route leaks.
\end{exercise}

\begin{exercise}
Design a stress-testing framework for a financial institution that includes cascading-counterparty scenarios.
\end{exercise}

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A team's power-grid analysis shows acceptable steady-state stability; the engineering question is the cascade resilience to specific contingencies. State the engineering approach.
\end{exercise}

\begin{exercise}
A team's supply chain has experienced two major disruptions in the past three years; the team is considering reshoring some production. State the engineering trade-offs.
\end{exercise}

\begin{exercise}
A team is implementing chaos engineering on a production microservice deployment; the engineering question is whether to inject failures that could cascade. State the engineering posture.
\end{exercise}

\begin{exercise}
A team's BGP-routing configuration includes a complex policy hierarchy; the team is unable to verify all the failure modes. State the engineering response.
\end{exercise}

\begin{exercise}
A team's banking customer is heavily exposed to a single counterparty; the team is considering whether to recommend diversification. State the engineering analysis.
\end{exercise}

\begin{exercise}
A team is asked to model the cascade risk of an autonomous-vehicle perception failure in a connected-vehicle network. State the engineering approach.
\end{exercise}

\begin{exercise}
A team's transportation network has a single critical bridge; the engineering question is whether to add a parallel route, expand capacity, or accept the risk. State the engineering analysis.
\end{exercise}
