\chapter{Human factors and operator error}
\label{vol10:ch09}

\epigraph{The operator did exactly what the system asked of them, and the system was wrong to ask it.}{}

\chapmeta{Half-life: durable. Archetypes: interface, ethics, failure. Exercise target: 20. Project track: hybrid.}

\noindent
"Operator error" is the most common attribution in accident reports and the most misleading. When an investigation traces an accident to "pilot error," "operator inattention," or "human factors," it has typically reached the proximate cause and stopped; the design, the procedure, the workload, the alert architecture, and the organisational pressure that produced the conditions under which the human's action was reasonable in the moment remain unexamined. This chapter is about the human as part of the engineered system: the cognitive architecture under which operators actually behave, the design choices that determine whether their behaviour is safe or catastrophic, and the cases that have become canonical in the human-factors literature.

\section{The human as part of the system}\pathtag{core}

The engineering discipline grows up alongside the assumption that the system is the hardware and the software, with the operator as an external input. The human-factors discipline grew up against that assumption. The operator is part of the system: a control element with sensing, processing, decision, and actuation; with documented failure modes, time constants, and operating envelopes; with capacity that depends on workload, training, time of day, fatigue, and the state of the engineered components they are interacting with. A system safety analysis that treats the operator as a black-box source of "correct action when needed" produces unsafe systems; a system safety analysis that treats the operator with the same engineering rigour as the hardware produces designs whose failure rates can be predicted and whose operator-action failure modes can be designed against.

The disciplines that have grown to do this work are several. \emph{Human factors engineering} is the design-oriented discipline: how should the controls be laid out, the displays formatted, the alerts annunciated, the procedures organised, so that the human's cognitive characteristics are supported rather than fought. \emph{Ergonomics} historically denotes the physical-anthropometric subset (reach envelopes, force limits, postural loading) but is often used synonymously with HF in European usage. \emph{Cognitive systems engineering} (Hollnagel and Woods) treats the human-machine system jointly: the unit of analysis is not the human alone or the machine alone but the cognitive system that emerges from their interaction. \emph{Resilience engineering} (Hollnagel; Dekker; Woods) treats the operator as the active source of adaptive response to the unforeseen, asking what makes operators capable of catching errors and absorbing surprises rather than what their failure rates are \cite{text:hollnagel-erik-2014}.

Across all of these, a few claims recur. Operators do not fail randomly; they fail in patterns that depend on the design of the system they are operating, on their training, on the workload they are under, and on the time available. The same operator who acts brilliantly under one workload acts disastrously under another. The same procedure that works in routine cases produces errors when applied to a non-routine case. The same alarm that catches the operator's attention when it is the only alarm fails to catch attention when it is one of forty simultaneous alarms. The patterns are predictable; the predictability is what makes human-factors engineering possible.

The dominant taxonomies of operator error came from Jens Rasmussen (1983) and James Reason (1990) \cite{paper:rasmussen-skills-rules-knowledge-1983,text:reason-human-error-1990}. Rasmussen's skills-rules-knowledge (SRK) framework distinguishes three levels of cognitive operation; Reason's slip-lapse-mistake taxonomy distinguishes three error mechanisms; Reason's later Swiss-cheese model maps how individual operator errors interact with latent organisational conditions to produce accidents (Chapter 1). These are the canonical frameworks that current accident investigations reach for; current operator-training programs are organised around them; current cockpit, control-room, and clinical interface design references them.

\begin{archetype}[interface]
Human factors is the interface archetype's purest expression in this volume. Two components, the human and the machine, communicate through an interface; the interface's properties (latency, fidelity, salience, ambiguity, completeness) determine whether the human's actions match the machine's state. An interface that is congruent with human cognition produces correct action under stress; an interface that fights human cognition produces error.
\end{archetype}

\section{Skill-based, rule-based, knowledge-based behaviour (Rasmussen)}\pathtag{core}

Rasmussen's SRK framework partitions cognitive operation into three levels distinguished by the kind of information being processed and the consciousness of processing \cite{paper:rasmussen-skills-rules-knowledge-1983}.

\textbf{Skill-based behaviour} is automatic, sensorimotor, executed without conscious attention. Driving a familiar route, typing a familiar word, adjusting a control knob to a familiar setting. The cognitive cost is low; the error mode is the slip (right intention, wrong action) or the lapse (right intention, action omitted). Skill-based behaviour is fast (response latencies in the hundreds of milliseconds) and energy-efficient; it is what frees attention for higher-level concerns. It fails when the situation changes in ways the automated action does not accommodate (the familiar route is closed; the familiar word's spelling is irregular; the control's familiar setting is wrong for the current state).

\textbf{Rule-based behaviour} is procedural, conscious but rapid: the situation is recognised as a member of a known class, and the response is the rule learned for that class. "If pressure exceeds 200, open the relief valve." "If altitude deviation alert sounds, check assigned altitude and correct." "If patient temperature exceeds 39.5, administer antipyretic and notify physician." Rule-based behaviour is the workhorse of trained operation; it is what the procedure manual codifies. The cognitive cost is moderate; the error mode is the misrule (wrong rule chosen for the situation, often because the situation is misclassified) or the rule misapplied (right rule, wrong execution).

\textbf{Knowledge-based behaviour} is deliberative, slow, effortful: the situation is novel, no learned rule applies, and the operator must reason from first principles using the mental model of the system. Diagnosing an unfamiliar fault. Planning around an unexpected obstacle. Constructing a response to a configuration the training did not cover. Knowledge-based behaviour is high-cost (response latencies in seconds to minutes; cognitive resources fully occupied) and high-error (the mental model may be incomplete or wrong; reasoning under time pressure produces shortcuts that mislead).

The framework's design implications are direct. The interface should support the operator's current level of operation: skill-based control should be supported with displays and controls whose physical layout maps to the action; rule-based operation should be supported with displays that make the situation's class unambiguous and procedures that match the situation; knowledge-based operation should be supported with displays that show the system's actual state and the affordances available, with diagnostic aids, and with workload reduction (alarm consolidation, automated routine actions, time afforded to deliberate). A common failure pattern in poorly-designed systems is that the operator is forced into knowledge-based operation by the design (the displays do not support classification; the procedures do not cover the situation), at the very moment when the situation is most stressful and least conducive to slow deliberation.

The SRK framework explains the canonical interface design errors. A control whose action depends on the system's state (a single button whose effect is "start" in one mode and "abort" in another) defeats skill-based execution because the same motor program produces opposite effects. An alarm display whose ten simultaneous alarms force the operator to read each in detail to determine which is the dominant condition defeats rule-based response because the situation's class is masked by noise. A diagnostic display that shows raw sensor readings without a system-state indication forces knowledge-based reasoning every time, even for routine faults.

\begin{principle}[Match interface to cognition]
Design interfaces to support skill-based operation for routine action, rule-based operation for recognisable conditions, and knowledge-based operation for novel situations, with explicit attention to which level the operator is in. An interface that forces the operator into a level the situation does not warrant either wastes cognitive resources (knowledge-based when rule-based would suffice) or produces error (rule-based when knowledge-based is required).
\end{principle}

\section{Cognitive biases under stress}\pathtag{standard}

The human cognitive architecture exhibits well-documented systematic deviations from optimal information processing. The biases are not failures of intelligence; they are the consequences of cognitive shortcuts (heuristics) that work well in routine cases and fail predictably in non-routine cases. Under stress (time pressure, high stakes, sleep deprivation, multiple simultaneous demands), the shortcuts dominate more strongly because the deliberate processes are inhibited.

\textbf{Confirmation bias.} Once a hypothesis is formed, the operator interprets ambiguous evidence as confirming the hypothesis and dismisses or rationalises disconfirming evidence. The pilot who believes the autopilot is engaged interprets the aircraft's behaviour as consistent with engagement; the operator who believes the relief valve has closed interprets rising pressure as a measurement anomaly; the controller who believes the aircraft is on assigned route interprets the radar deviation as an instrument glitch. Confirmation bias accelerates routine processing (the routine hypothesis is usually correct) and fails when the situation is anomalous (the unusual hypothesis is dismissed too quickly).

\textbf{Anchoring.} The first hypothesis considered or the first value seen sets a reference that subsequent estimates do not move far from. Initial readings of an instrument bias subsequent estimates of the true value; initial diagnoses bias subsequent differential consideration. Anchoring is harmful in diagnosis because the actual condition may differ substantially from the initial impression and the deliberation needed to recognise that difference is suppressed.

\textbf{Availability.} The probability of a class of event is estimated by how easily examples of the class come to mind. Recent events, vivid events, and events frequently discussed are estimated as more probable than they are; rare events are estimated as less probable. The pilot who has not seen a runaway-stabiliser event in 10,000 hours estimates the probability as zero; the operator who has seen three control-system glitches this week estimates the probability of a fourth as high. Availability bias produces systematic miscalibration of low-frequency, high-consequence events.

\textbf{Hindsight bias.} After an outcome is known, the prior probability of that outcome is misremembered as having been higher than it was. The investigator who knows the accident occurred reads the operator's pre-accident decisions as obviously wrong; the operator at the time may have had no available evidence pointing to the actual outcome. Hindsight bias is the cognitive mechanism behind much of the "obvious in retrospect" framing of accident reports and is the reason serious accident analysis requires investigators to reconstruct the operator's perspective as it actually was, not as the post-hoc view makes it appear.

\textbf{Loss aversion and framing.} Decisions involving uncertain gains and losses are made differently depending on whether the choices are framed as gains or losses; losses are weighted more heavily than gains of equivalent magnitude \cite{paper:kahneman-tversky-1979}. Under operational pressure, an option framed as "avoiding a known cost" is preferred over an option framed as "accepting a possible gain," even when the expected utilities favour the second. The pilot who accepts a marginal approach to avoid the cost of a go-around is choosing under loss-aversion framing; the same pilot in a "gain" frame (a successful landing as benefit) would weigh the choice differently.

\textbf{Plan continuation.} An ongoing plan is continued past the point at which evidence suggests it should be revised; the cognitive cost of revising and replanning is high, and the partial commitment that the plan represents biases decision against abandonment. "Get-there-itis" in aviation; "summit fever" in mountaineering; "deliver-by-deadline" in software releases. Plan continuation is a fast-paced version of the same effect that anchoring produces in diagnosis.

\textbf{Workload-induced narrowing.} Under high workload, attention narrows to the immediate task; peripheral signals are missed; alternative interpretations are not considered. The narrowing is adaptive (it focuses cognitive resources on the immediate problem) and dangerous (the missed signal may be the one that warns of a different problem). Eastern 401 \cite{acc:ntsb-aar73-14} is the canonical case: workload around the landing-gear indicator narrowed the crew's attention so completely that the altitude-deviation aural alert was not noticed.

The design responses to these biases are constraints on the human as a decision-maker. Procedures that require explicit consideration of alternative hypotheses (differential diagnosis as a habit; "what could I be missing?" as a callout in cockpit CRM). Displays that present absolute values rather than relative anchors. Decision aids that flag candidate hypotheses the operator may not have considered. Forcing functions that require explicit acknowledgement before continuing past decision points. Training that exposes operators to the patterns of their own biases so the bias is named and recognisable when it is happening.

\section{Mode confusion and automation surprises revisited from Volume IX}\pathtag{core}

Volume IX Chapter 14 introduced mode confusion: the operator's mental model of the automation's current state diverges from the automation's actual state, and operator action that would be correct in the believed mode is incorrect in the actual mode. The phenomenon is the central failure mode in highly-automated systems; this chapter expands the treatment as an HF failure type with its own dynamics.

Modern automated systems (cockpit autopilots, autothrottles, autoland systems; industrial process control; medical infusion pumps; automotive driver assistance) operate in many modes, with mode transitions that may be commanded by the operator or initiated by the automation in response to system state. The number of modes can be large (a modern cockpit may have dozens of autoflight modes, with nested sub-modes); the transitions between them can be complex (some modes are entered automatically when conditions are met, exited automatically when conditions cease; some require explicit operator entry; some forbid certain transitions). The operator's job is partly to remain aware of which mode the automation is currently in, what that mode commands, and what the consequences of operator actions will be in that mode. Mode awareness is cognitively expensive; under workload, it is the first thing to fail.

The canonical failure modes:

\textbf{Mode awareness lapse.} The operator believes the automation is in mode A; it is actually in mode B. Operator commands appropriate to mode A produce mode-B effects. The Air France 447 sequence \cite{acc:bea-af447}, after the loss of airspeed indications caused the autopilot to disengage and the flight control system to revert from "normal law" (which prevents the aircraft from being stalled) to "alternate law" (which does not), is structurally a mode-awareness lapse: the crew's stick inputs that would have been bounded by normal law's protections were not bounded by alternate law, and the resulting nose-up command stalled the aircraft.

\textbf{Mode transition without annunciation.} The automation changes modes due to a triggering condition; the annunciation is absent, transient, or peripheral; the operator does not realise the mode has changed. Eastern 401's autopilot transitioned out of altitude-hold without the crew noticing \cite{acc:ntsb-aar73-14}; the change in mode was not salient against the workload-narrowed attention.

\textbf{Mode trap.} The operator's intended command requires the automation to be in a particular mode, but the operator does not realise which mode is required and commands instead a sequence that the current mode does not interpret as intended. The pilot who commands "altitude 5000" expecting the autopilot to climb to and capture 5000, but whose current mode is one in which the altitude entry is only a reference and the aircraft does not climb without an additional command, is in a mode trap.

\textbf{Inconsistent mode interfaces.} The same physical control has different effects in different modes; the same display element has different meaning in different modes. The cognitive cost of remembering the mode-specific mapping increases with the number of modes; under workload, mappings collapse to the most-recent or most-frequent, with the wrong action commanded in the off-frequency case.

\textbf{Automation complacency.} The operator, having seen the automation perform correctly in most cases, defers to its action without monitoring; when the automation reaches a state it is not designed for, the operator is not actively engaged and does not catch the error promptly. The phenomenon is well-documented in autoland systems, in adaptive cruise control, in algorithmic trading, and in clinical decision support; it is the cognitive consequence of automation that is reliable enough to be trusted but not reliable enough to be trustworthy without supervision.

The design responses are: minimise the mode count consistent with function; make mode transitions visible (the annunciation must be impossible to miss, not merely present); make mode actions consistent across modes where possible; provide direct readout of the automation's current behaviour, not just its mode label; train operators on the mode taxonomy and on the recognised failure modes of mode confusion; require the operator to be actively engaged at appropriate intervals to defeat complacency.

\section{Crew resource management}\pathtag{core}

Crew resource management (CRM) is the training discipline developed in commercial aviation in the late 1970s and early 1980s in response to Tenerife \cite{acc:ciaiac-tenerife-1978}, Eastern 401 \cite{acc:ntsb-aar73-14}, and a cluster of other accidents in which technically skilled crews failed because of failures in coordination, communication, and decision-making rather than technical skill. The 1979 NASA workshop "Resource Management on the Flight Deck" was the discipline's foundational event; the United Airlines training program of 1981 was the first major airline mandate; the FAA's Advisory Circular 120-51 codified the discipline; international adoption followed in the 1980s and 1990s; CRM is now mandated by every major civil aviation authority and by ICAO.

The discipline's central claims:

\textbf{Communication is a skill, not a personality trait.} Effective cockpit communication requires explicit techniques: standard phraseology (which forbids ambiguous formulations like Tenerife's "we are now at takeoff"); structured callouts at decision points; closed-loop confirmation (the receiver repeats the message; the sender confirms the repetition); explicit verbal queries and explicit verbal acknowledgements. These techniques are trained and tested; they are not assumed to be present in trained pilots.

\textbf{Authority gradient must be calibrated.} An overly steep authority gradient (the captain is unchallengeable; the first officer cannot effectively raise concerns) produces accidents in which a captain's error proceeds uncorrected; an overly shallow gradient (no clear command structure) produces accidents in which no one is in charge of the integrated picture. CRM trains both halves: captains receive training in inviting and receiving challenges; first officers receive training in raising concerns in language and at times the captain can act on.

\textbf{Workload management is a crew skill.} Tasks should be allocated explicitly between crew members; when one crew member is heads-down on a problem, another is explicitly responsible for flying; transitions in who is flying are verbal and confirmed; workload-shedding (which tasks can be deferred; which can be delegated to automation; which can be handed off to other resources) is a planned response to high-workload situations rather than an emergent improvisation.

\textbf{Situation awareness is a crew product.} The crew's shared mental model of the aircraft state, the environment, the plan, and the threats is built through explicit verbal sharing; private mental models that diverge between crew members produce coordination failures. CRM techniques include explicit briefings at transitions (top-of-descent briefing; approach briefing; missed-approach briefing) that synchronise the crew's models.

\textbf{Decision-making is a structured process.} CRM teaches structured decision frameworks (FOR-DEC, T-DODAR, and similar) that decompose decisions into Facts, Options, Risks/benefits, Decision, Execution, Check. The structure forces explicit consideration of alternatives and explicit assignment of action and review.

CRM's empirical track record is the post-1980s decline in human-factors causation of major accidents in commercial aviation; while many factors contributed (better automation, better training, better procedures, better regulation), CRM is widely credited as a substantial contributor. The discipline has been adapted for surgical operating rooms (sometimes "team resource management" in healthcare; the 1999 Institute of Medicine report \emph{To Err Is Human} popularised the framing in medicine), for nuclear control rooms, for offshore drilling, for spaceflight mission control, and for incident command in emergency response. The training is now sufficiently transferable that "CRM-style" training is recognisable across domains.

\section{The myth of operator error and what it usually hides}\pathtag{core}

The single most-corrected mistake in accident investigation is the attribution of an accident to operator error. The attribution is rarely false (an operator did act, and the action did contribute), but it is rarely the whole story, and the investigations that stop at the operator's action systematically miss the design, procedural, organisational, and regulatory conditions that produced the action. Dekker's work, drawing on the resilience-engineering tradition, frames the issue as the choice between "old view" (operator-error attribution) and "new view" (the operator's action was a reasonable response to the system as presented; the question is why the system presented it that way) \cite{text:dekker2014}.

The diagnostic moves that distinguish the new view from the old:

\textbf{Why was the operator's action locally reasonable?} An accident-causing action was almost always reasonable in the moment, given what the operator knew, what the displays showed, what the procedures called for, and what the workload allowed. The first analytical move is to reconstruct the operator's perspective at the time, without the post-hoc knowledge of the outcome. The Air France 447 first officer's nose-up sidestick input was reasonable under his (incorrect) belief that the aircraft was in a low-altitude regime requiring nose-up input; the reason for the belief, the absent or misinterpreted indications that would have corrected it, and the design choices that allowed the misinterpretation are the substantive analysis \cite{acc:bea-af447}.

\textbf{What in the design allowed this action to produce this outcome?} An operator action that produces a catastrophic outcome is doing so partly because the system permits the action, fails to warn about the action, or makes the action's consequences invisible. Design changes that close these channels reduce the consequences of the operator's error class. The MCAS redesign in the post-737-MAX response (Chapter 8) is the textbook example: the design changes ensured that the operator action class that produced the accident could not produce the same outcome in the redesigned system.

\textbf{What in the procedure or training would have predicted this action?} An operator action is more likely if it is what training teaches, what procedures direct, or what informal practice expects. An accident often reveals that the formal training, the formal procedure, and the informal practice diverge; the operator who follows informal practice rather than formal procedure is responding to incentives the organisation has created.

\textbf{What in the workload, fatigue, or organisational pressure made this error more probable?} Operators do not act with constant capacity; their capacity is modulated by workload (concurrent demands, time pressure, alarm flood), fatigue (sleep deprivation, circadian phase, time on task), and pressure (production goals, schedule, organisational culture around speed or cost). An accident that occurred under high workload, fatigue, or pressure was made more probable by those conditions; the conditions themselves are organisational and design choices, not operator characteristics.

\textbf{Has this class of operator error occurred before, and if so, why was it not designed against?} The accident under investigation is often one of many similar events; the difference between an incident (no consequence) and an accident (severe consequence) is luck and timing rather than the underlying operator action. The chapter on near-misses (Chapter 13) develops this point; the present point is that operator-error attribution loses its persuasive force when the same operator-error class has been observed many times before without producing change.

The new view is not the absence of accountability. Operators can be reckless, can be impaired, can act in violation of procedure or training; these cases exist and warrant accountability. The new view is the demand that accountability be assigned correctly: the operator who could have acted differently in a system that supported correct action is responsible; the operator whose system did not support correct action under the actual conditions of the event is not the substantive cause of the failure even if their action was the proximate cause. The distinction is what separates retributive investigation (find someone to blame; close the case) from constructive investigation (find what to change; reduce the next accident's probability).

\section{Case: Air France 447, Eastern 401, Tenerife}\pathtag{core}

\subsection*{Tenerife, 27 March 1977}

The Tenerife collision killed 583 people when two Boeing 747s collided on the runway at Los Rodeos Airport in dense fog \cite{acc:ciaiac-tenerife-1978}. The KLM 747 began its takeoff roll without takeoff clearance while the Pan Am 747 was still backtracking on the same runway. The Spanish investigation identified the proximate sequence: KLM received an ATC route clearance; the KLM captain advanced throttles; the first officer transmitted "we are now at takeoff" or "we are now uh taking off," a non-standard phraseology whose meaning was ambiguous; the tower responded "OK $\ldots$ stand by for takeoff, I will call you," a transmission that overlapped with a Pan Am transmission asking whether the runway was clear; the simultaneous transmissions produced a heterodyne squeal in the KLM cockpit; the KLM crew heard only the "OK" portion of the tower's response and did not hear the Pan Am transmission at all; KLM continued the takeoff roll; the collision occurred when KLM had achieved rotation speed and was airborne by a few meters.

The human-factors lessons aggregate to a CRM curriculum. The non-standard phraseology ("we are now at takeoff") was ambiguous between intent and action; standard phraseology after Tenerife forbids any use of "takeoff" except in the takeoff clearance itself ("cleared for takeoff"). The senior KLM captain (head of the airline's 747 training department) created an authority structure in which the first officer's voiced concerns were not pressed; CRM training is the institutional response that gives first officers explicit language and protocol for challenging the captain's decisions. The radio congestion that produced the heterodyne squeal was a failure of the communication channel itself; subsequent ATC procedures and equipment changes reduced (but did not eliminate) the channel's susceptibility to this failure mode. The commercial pressure around crew duty time and refuelling time was a background condition that shaped the captain's decision speed; the organisational response is the codification of duty-time limits that are enforced regardless of operational pressure.

\subsection*{Eastern Air Lines Flight 401, 29 December 1972}

Eastern 401 crashed in the Florida Everglades killing 101 of 176 on board \cite{acc:ntsb-aar73-14}. The Lockheed L-1011 had been holding west of Miami at 2000 ft while the crew troubleshot a burnt-out nose-landing-gear position indicator bulb. During the troubleshooting, the autopilot's altitude-hold mode disengaged (most likely from inadvertent pressure on the control column), and the aircraft began a gradual descent at approximately 200 feet per minute. The altitude-deviation aural alert sounded at the second officer's station while the second officer was in the avionics bay below the cockpit attempting to visually confirm gear position; the alert was not noticed by the remaining cockpit crew. Air traffic control noticed the deviation on radar but queried with insufficient specificity to alert the crew. The aircraft descended into the swamp 80 seconds after the alert, with the first verbal indication in the cockpit being the first officer's "we did something to the altitude" seconds before impact.

The human-factors lessons are foundational. Task-fixation: a low-cost indicator bulb produced complete attentional capture in three highly-trained pilots, illustrating how routine low-stakes troubleshooting can absorb the cognitive resources that a routine high-stakes monitoring task requires. Task allocation: the captain had not designated who was flying the aircraft while the others worked the indicator; the implicit assumption that "the autopilot is flying" placed the autopilot in the role of a crew member without any of the monitoring crew members assigned to oversee it. Alert placement: the altitude-deviation aural alert sounded at the second officer's station; the second officer was in the avionics bay; the alert was therefore inaudible to the remaining crew. The post-Eastern-401 response duplicated such alerts at all crew stations and tightened the protocols for designating the pilot flying. The accident's foundational role in CRM development is documented in the post-1979 NASA workshop literature.

\subsection*{Air France 447, 1 June 2009}

AF447, an Airbus A330-203 (registration F-GZCP), crashed into the Atlantic Ocean during a flight from Rio de Janeiro to Paris on 1 June 2009, killing all 228 occupants \cite{acc:bea-af447}. The BEA investigation identified the sequence: the aircraft entered an area of convective weather at cruise altitude (FL350); the pitot tubes iced over, producing inconsistent and unreliable airspeed indications; the autopilot disengaged automatically (it cannot operate without reliable airspeed); the flight control system reverted from "normal law" (which has hard envelope protections preventing the aircraft from being stalled) to "alternate law" (which does not). The first officer flying made nose-up sidestick inputs that were not corrected by the protections that would have applied in normal law; the aircraft entered a high-altitude stall; the crew did not recognise the stall (the stall-warning aural alert had stopped sounding because the airspeed indications fell below the threshold at which the warning is enabled; the warning resumed when the crew briefly pitched down enough to increase airspeed, then stopped again when they returned to nose-up); the aircraft descended from FL350 to the ocean in approximately 3 minutes 30 seconds, with the crew never recognising the stall they were in.

The human-factors lessons are dense. Mode awareness: the transition from normal law to alternate law removed protections that the first officer's training and intuition had been built around; the BEA report identifies the inadequate annunciation of the mode change as a contributing factor. Stall recognition: the modern airliner's flight envelope protections in normal law mean that crews never train stall recognition at the cruise altitudes where AF447's stall occurred; the high-altitude stall regime was outside the experience and training of the crew. The stall-warning logic that disabled the warning when airspeed indications fell below threshold produced perverse feedback (pitching down to recover, the correct action, restarted the warning; pulling up, the wrong action, silenced it). Crew coordination: the captain was on a rest break when the autopilot disengaged; he returned to the cockpit but was unable in the available time to assemble a clear picture of what was happening and the recovery action that was needed; the two first officers' actions were not effectively coordinated.

The BEA report's design and training recommendations centred on the gap between the modern automated airliner's normal-operations regime (where the autopilot and the protections do almost all of the work) and its degraded-mode regime (where the protections are absent and the manual-flying skill required is the same as in earlier aircraft generations, except never practiced). The recommendations propelled industry-wide changes in cruise-altitude manual-handling training, in stall-recognition training for high-altitude regimes, in the design of mode-transition annunciations, and in the logic of stall-warning systems.

\begin{principle}[Investigate to the design, not to the operator]
Stop the investigation at "operator error" and the next accident in the same class will reproduce the same outcome. Trace the operator's action to the design, the procedure, the training, the workload, and the organisational pressure that produced it, and the design changes that reduce the action class's consequences become identifiable. The discipline is the discipline of the new-view investigation; the alternative is the production of repeating accidents that no one has the leverage to change.
\end{principle}

\begin{project}[CVR transcript study]
Obtain the cockpit-voice-recorder transcript of one accident with a substantial human-factors causal contribution (Tenerife, Eastern 401, Air France 447, Asiana 214, or another from the public CVR-transcript record). Read the transcript end to end. Mark every point at which an alternative intervention by any crew member could plausibly have changed the outcome; for each, identify the cognitive, procedural, or design factor that prevented or shaped the intervention that actually occurred. Write a four-page analysis: a timeline; a list of intervention points with mechanisms; a synthesis of which factors recur across the points; one design or procedure change that would have most reduced the accident's probability. Deliverable: the four-page document.

\textbf{Hazard.} The exercise is emotionally difficult; the transcripts capture the last moments of crews whose families have read these documents. The point is to learn from those crews and from the people who investigated the accidents; the analytic register is the same one the BEA, NTSB, and AAIB use in their reports. Do not write in a tone that the families would find disrespectful; the engineering register is the appropriate register.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
The altitude-deviation aural alert on Eastern 401 sounded when the aircraft descended 250 feet below the assigned 2000 ft altitude \cite{acc:ntsb-aar73-14}. The aircraft was descending at approximately 200 feet per minute. Compute the time from the alert sounding to impact.
\end{exercise}

\begin{exercise}
A pilot's response to an unexpected aural alert in cruise has a median latency of approximately 4 seconds (reaction to alert), plus 3 seconds (recognition of meaning), plus 5 seconds (response decision), plus 2 seconds (response execution). Compute the total median response time. If the alert is one of N simultaneous alerts, the recognition time scales as $3 + 1.5(N-1)$ seconds (a heuristic for alarm-flood interpretation). Compute response time for $N = 1, 5, 10, 20, 40$.
\end{exercise}

\begin{exercise}
The Tenerife heterodyne occurred when KLM's reply to the tower overlapped with Pan Am's transmission \cite{acc:ciaiac-tenerife-1978}. If the typical radio transmission is 4 seconds and the controller responds 1 second after the first transmission ends, estimate the probability per minute that two simultaneous transmissions occur in a setting with 5 aircraft on the same frequency each transmitting on average twice per minute.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the predicted operator response time as a function of cognitive load using the simplest one-channel model: a queue of pending cognitive tasks each requiring a service time $\tau$, arriving at rate $\lambda$. Identify the load regime in which an additional task increases median response time substantially.
\end{exercise}

\begin{exercise}
For a two-pilot cockpit with explicit roles (one pilot flying, one pilot monitoring), derive the conditions under which a fault detection improves over single-pilot operation. Identify the parameters (independent detection probability, communication latency, correction latency) on which the improvement depends. Identify the conditions under which the two-pilot architecture does not improve over single-pilot (when communication is degraded, when monitoring is not active, when authority structure blocks the monitor's correction).
\end{exercise}

\begin{exercise}
Derive Reason's Swiss-cheese model as a probabilistic statement: given $n$ defensive layers each with a probability $p_i$ of containing a "hole" through which the hazard can pass, the probability of an accident is the probability that the holes align for the specific event. Identify the conditions under which the model predicts the probability accurately and the conditions (correlated holes; latent shared causes) under which it underestimates.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the fraction of commercial-aviation accidents in the period 1970-1990 attributed primarily to "crew/human-factors" causation. Compare to the fraction in the period 1990-2010 (post-CRM-adoption). Identify the magnitude of CRM's plausible contribution.
\end{exercise}

\begin{exercise}
Estimate the probability that an operator under workload of 3 simultaneous tasks notices a peripheral aural alert located at a station the operator is not currently attending. Compare to the probability of noticing the alert located at the operator's current station. Identify the design implication.
\end{exercise}

\begin{exercise}
Estimate the alarm rate per hour in a typical modern process-plant control room. Identify the fraction of alarms that are actionable and the fraction that are "alarm-flood noise." Identify the design implication for alarm prioritisation and consolidation.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Implement a simulation of a two-pilot cockpit in which an autopilot mode change occurs with probability $p_m$ per hour, mode-change annunciations are noticed with probability $p_a$ given crew workload $W$, and a missed mode change produces accident probability $p_c$ if not corrected within time $T$. Sweep $W$ and identify the workload regime in which the system's accident probability rises sharply.
\end{exercise}

\begin{exercise}
Simulate the alarm-flood phenomenon: during a process upset, simultaneous alarms arrive at rate $\Lambda = 10$ per minute and the operator's recognition rate is $\mu = 6$ per minute. Plot the operator's alarm-recognition queue depth over a 10-minute upset and compute the fraction of alarms that are missed or delayed past their actionable window.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design an aural-alert architecture for a three-station cockpit that ensures every critical alert is heard at every station, while avoiding alarm flood. Specify the alert tones, their priorities, the architecture of their distribution, and the suppression rules.
\end{exercise}

\begin{exercise}
Design a CRM training module for surgical operating-room teams that adapts cockpit CRM concepts to the surgical setting. Identify the equivalent of "standard phraseology," "authority gradient calibration," "workload management," and "structured decision-making" in the surgical context.
\end{exercise}

\begin{exercise}
Design the post-accident response for a hypothetical airline whose recent accident closely resembles Eastern 401 \cite{acc:ntsb-aar73-14}. Specify the technical change, the procedural change, the training change, and the organisational change recommended. Justify each.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
An accident investigation has identified the proximate cause as "the pilot pulled back on the control column when the situation called for nose down." The investigation has not asked why the pilot believed nose-up was correct. Critique the investigation and identify what additional analytical moves are required.
\end{exercise}

\begin{exercise}
Argue: the design changes that would have prevented Air France 447 \cite{acc:bea-af447} were technically feasible at the time of the aircraft's certification and were not implemented because of organisational rather than technical constraints. Identify the design changes; identify the organisational constraints; evaluate the argument.
\end{exercise}

\begin{exercise}
The Tenerife KLM captain was the head of the airline's 747 training department \cite{acc:ciaiac-tenerife-1978}. Discuss how CRM training, had it existed at the time, would have changed the cockpit-coordination dynamic in the moments before takeoff. Identify the specific CRM technique that would have been most consequential.
\end{exercise}

\begin{exercise}
A workplace investigation attributes a manufacturing-floor injury to "the operator failed to engage the safety lockout." The operator says the lockout was inoperative for the past two weeks and had been reported to maintenance. Apply the new-view diagnostic moves and identify what the investigation should examine next.
\end{exercise}

\begin{exercise}
A medical infusion-pump user interface allows the operator to enter a drug dose in mg/kg/hr but defaults to the previous patient's value when a new patient is selected. A nurse, working at the end of a 12-hour shift with three concurrent admissions, enters a new patient and overlooks the inherited value, administering the previous patient's dose to the new patient. Using the SRK framework \cite{paper:rasmussen-skills-rules-knowledge-1983} and the cognitive-bias catalogue from this chapter, identify which mechanisms made the error probable. Identify the design changes (interface, defaults, alerts, forcing functions) that would have most reduced the probability.
\end{exercise}

\begin{exercise}
A car's adaptive-cruise-control system can be in any of four modes: off, set-but-not-engaged, engaged-following-vehicle, engaged-no-vehicle-ahead. The HUD indicates the current mode by a small icon at the bottom right of the speedometer display. The driver, accustomed to engaged-following-vehicle behaviour (which slows the car when the lead vehicle slows), is surprised when the car does not slow as the lead vehicle merges out of lane; the actual mode at that moment is engaged-no-vehicle-ahead, which maintains the set speed without responding to other traffic. Categorise the failure mode using the mode-confusion taxonomy in this chapter. Identify three specific design changes that would reduce the probability of recurrence.
\end{exercise}
