\chapter{Conceptual design and sketching}
\label{vol11:ch04}

\epigraph{Generate ten concepts before committing to one; commit to the right one rather than the first.}{}

\chapmeta{Half-life: medium-life. Archetypes: optimisation, balance. Exercise target: 20. Project track: hybrid.}

\noindent
Conceptual design is the stage between requirements and detailed design at which the engineer (or team) generates candidate solutions and selects among them. The stage is the highest-leverage point in the design process: decisions made here constrain the rest of the design; concepts that are not generated at this stage cannot be selected later; concepts that are committed prematurely shape the detailed design's space in ways that are difficult to reverse. This chapter develops the discipline of generation, the discipline of selection, and the failure modes (premature commitment, weak generation, biased selection) that produce designs that are competent within their constraints but that have surrendered the better alternatives at the conceptual stage.

\section{Generation of alternatives}\pathtag{core}

The substantive content of conceptual design is the generation of multiple alternatives. A single concept is not a concept; it is an assumption. The discipline insists on breadth: multiple concepts explored before any one is committed to. The reasoning:

\textbf{The first concept is rarely the best.} The first concept that comes to mind is shaped by recent experience, by familiarity, by the path of least cognitive resistance. The first concept may be adequate; it is rarely best. The discipline of generating ten concepts (or however many is appropriate) before selecting one produces a better-than-first-concept solution.

\textbf{Selection requires a choice set.} A concept cannot be selected without alternatives to compare it against. Selecting among one option is committing to that option; the "selection" is conceptual cover for what was already a commitment.

\textbf{Generation surfaces the design space.} The process of generating multiple concepts surfaces the dimensions along which concepts can vary, the constraints that bind, the assumptions that the team has been making. The generation activity is itself epistemically productive; concepts that are not selected still contribute to the team's understanding of the problem.

\textbf{Diverse concepts hedge against unforeseen issues.} A single concept that fails at detailed design (because of a manufacturing constraint, a cost overrun, a regulatory issue) leaves the project with no fallback; multiple concepts that have been explored provide alternatives if the first choice does not survive.

The generation techniques:

\textbf{Brainstorming.} The team generates ideas without immediate evaluation; the volume of ideas is the immediate goal; evaluation is deferred to a separate session. The classical rules (defer judgment; aim for quantity; build on others' ideas; welcome unusual ideas) are designed to produce a generative atmosphere. The technique is widely used; its effectiveness depends on facilitation discipline and on team composition.

\textbf{Brainwriting.} Each team member generates ideas individually in writing; the written ideas are then shared and built on. The technique produces more total ideas than verbal brainstorming and reduces the dominance of vocal participants; it works well for teams in which power dynamics suppress verbal participation.

\textbf{Worst-possible-idea.} The team is asked to generate the worst possible solutions to the problem; the discipline reverses the censoring instinct that often suppresses unconventional ideas. The worst-idea generation often surfaces the structural assumptions of the problem (why is X bad?) which subsequent reversal can convert into good ideas.

\textbf{Analogical reasoning.} The team identifies analogous problems in different domains (a biological system; a different industry; a different scale) and adapts the solutions. The technique opens the design space beyond the team's immediate domain; the discipline includes judgment about which features of the analogue transfer and which do not.

\textbf{First-principles generation.} The team derives candidate concepts from the underlying physics, chemistry, or mechanics of the problem rather than from prior solutions. The technique produces concepts that prior-solution-anchored generation would not; it requires deep domain knowledge.

\textbf{Systematic methods.} TRIZ (treated in Section 4 below), morphological analysis (Section 3), and similar systematic techniques produce concepts through structured exploration of the design space rather than through associative generation.

The volume of generation matters. The classical guidance is that ten concepts is a reasonable starting target for non-trivial problems; the team's commitment to generating ten (rather than two or three) forces the team past the obvious and into the less obvious. The discipline is the discipline of staying in generative mode for long enough to surface the non-obvious alternatives.

\begin{principle}[Generate before selecting]
The single concept that comes to mind first is not the design. The discipline of conceptual design is the discipline of generating multiple concepts before selecting any one, and selecting through structured evaluation against the requirements rather than through commitment to the first plausible idea. Designs that proceed from a single concept have skipped the highest-leverage decision the design process offers.
\end{principle}

\section{Sketching as engineering tool}\pathtag{standard}

Sketching is the engineering activity of representing concepts in low-fidelity visual form. The sketch is not a drawing in the artistic sense; it is a thinking tool. The engineer sketches to think, to communicate, to compare, to commit.

The functions sketching serves in conceptual design:

\textbf{Thinking by drawing.} The act of sketching surfaces design details that purely verbal or purely mental consideration misses. The sketch's spatial constraints (does this part actually fit where the concept requires?) surface dimensional issues; the sketch's representational decisions (how are the parts connected? what direction does the load flow?) surface structural and functional issues. The engineer sketches not to communicate a complete concept but to discover the concept's details through the sketch's medium.

\textbf{Communicating concepts within a team.} A sketch shared among team members carries information that verbal description cannot match. The team's discussion of the sketch is more productive than discussion of an undrawn concept; the sketch makes the concept concrete enough to react to.

\textbf{Comparing concepts.} A page of sketched concepts can be compared at a glance; the same concepts in text would require sequential reading and re-reading. The visual comparison surfaces relative strengths and weaknesses that text comparison would miss.

\textbf{Committing to detail-level decisions.} As the conceptual stage progresses, the sketches develop from rough thumbnails to more detailed representations. The progression is the engineer's commitment trajectory: each iteration commits to more specific decisions and rules out alternatives that the earlier sketches were ambiguous about.

The sketch is low-fidelity by intent. High-fidelity (CAD-quality) representations at the conceptual stage are counterproductive: they take too long to produce (reducing the number of concepts the team can explore); they signal more commitment than the conceptual stage warrants (making the team reluctant to discard the concept); they constrain the team's thinking to the precision the tool requires. The conceptual sketch is rough; the detailed CAD work comes later (Chapter 5).

The sketching discipline includes:

\textbf{Speed.} The sketch should take minutes, not hours. The investment per sketch is small so the team can generate many.

\textbf{Multiple views.} A concept often needs multiple views (plan, elevation, section, isometric) to capture its three-dimensional structure. The sketches are produced as needed.

\textbf{Annotation.} The sketch is annotated with dimensions (approximate), notes (the design intent; the open questions), and references (the requirements the sketch addresses).

\textbf{Iteration on the same sketch.} The sketch is modified as the concept develops; revisions are tracked through dated copies or through annotation. The history is part of the design record.

\textbf{Mediums.} Paper and pencil; whiteboards; tablet-and-stylus; printed CAD scaffold with hand annotation. The medium is chosen for the activity; paper is dominant for individual thinking; whiteboards for team discussion; tablets for documents that need to be preserved digitally.

The hand-sketching skill is part of the engineering toolkit. Engineering programs that teach sketching as a working discipline (rather than as an art-class activity) produce engineers who think more effectively at the conceptual stage. The settled editorial decision is to treat sketching as engineering practice, not as an optional add-on.

\section{Morphological analysis}\pathtag{standard}

Morphological analysis (also called the "morphological box" or "Zwicky box," after Fritz Zwicky who developed the technique in the 1940s) is a systematic method for generating concepts by combining choices across independent design dimensions.

The procedure:

\textbf{Step 1: Decompose the design problem into functional dimensions.} For each function the design must perform, identify the dimension. For a hand-held electric drill: power source; motor type; transmission; chuck mechanism; user interface; housing; cord/cordless. The functions are the rows of the morphological matrix.

\textbf{Step 2: Enumerate options for each dimension.} For each function, list the candidate solutions. For "power source": mains, lithium-ion battery, NiMH battery, pneumatic, hydraulic, hand-cranked. For each function, list as many candidate solutions as plausible. The options are the cells of the matrix.

\textbf{Step 3: Combine across dimensions to generate concepts.} A concept is a choice of one option per dimension. The total concept space is the product of the dimensions' option counts; for a drill with 6 dimensions each with 5 options, the space is $5^6 = 15{,}625$ concepts. The space is too large to enumerate; the technique is to sample it systematically.

\textbf{Step 4: Filter for feasibility.} Many combinations are incompatible (a pneumatic drill with a battery doesn't make sense). The infeasible combinations are eliminated; the feasible combinations are the candidates for further evaluation.

\textbf{Step 5: Identify promising regions.} The feasible space is clustered into regions sharing common characteristics; the regions are evaluated against the requirements to identify the most promising. Detailed concepts are then developed for the most promising regions.

The technique's value:

\textbf{Forces breadth in generation.} The engineer who would naturally have generated 3-5 concepts is forced by the technique to consider many more. The non-obvious combinations often surface concepts the engineer would not have generated through associative thinking.

\textbf{Surfaces the design's dimensions.} The discipline of decomposing the design into dimensions clarifies what the design's degrees of freedom are. The dimensions become explicit and the team's discussion can be structured around them.

\textbf{Supports systematic evaluation.} The feasibility filter is structured; the evaluation is across explicit dimensions; the trade-offs are visible.

The technique's limitations:

\textbf{Combinatorial explosion.} Large design spaces produce more combinations than can be evaluated. The technique requires judgment about which dimensions and options to include.

\textbf{Decomposition error.} The decomposition into dimensions may not be the natural decomposition for the problem; the technique's effectiveness depends on the decomposition's quality.

\textbf{Independence assumption.} The technique assumes the dimensions are independent (any option in one dimension is compatible with any option in another). Real designs have strong dependencies; the feasibility filter addresses some but not all.

The technique is widely used in product design, in chemical-process design, and in systems-engineering trade studies. The discipline of using it well is the discipline of decomposing thoughtfully, enumerating broadly, and filtering rigorously.

\section{TRIZ engineering-first}\pathtag{standard}

TRIZ (Theory of Inventive Problem Solving, Genrich Altshuller, 1946-1985) is a systematic method for engineering innovation derived from the analysis of patents. Altshuller and his collaborators reviewed approximately 200{,}000 patents and identified patterns of inventive solutions: recurring contradictions that engineers face, and recurring principles by which the contradictions are resolved.

The TRIZ framework's core elements:

\textbf{Contradictions.} An engineering problem typically arises from a contradiction between two requirements. Strengthening a structural member by adding material increases its weight, which is undesired. The contradiction is between "strong" and "light"; the conventional design move is to compromise (a structure that is strong enough and light enough); the inventive move is to resolve the contradiction (a structure that is strong and light, often by changing the structure's principle of operation).

\textbf{Contradiction matrix.} The TRIZ contradiction matrix lists 39 engineering parameters (weight, strength, speed, temperature, etc.) along both axes. Each cell identifies the parameter that improves (row) and the parameter that worsens (column); the cell's contents are the inventive principles that have been used to resolve that contradiction in the patent literature.

\textbf{40 inventive principles.} The inventive principles abstracted from the patent analysis: segmentation (divide the object into parts); taking out (separate the interfering part); local quality (use different material or properties for different parts); asymmetry (replace symmetric form with asymmetric); merging (bring closer in space or time); universality (combine multiple functions); nesting (place one inside another); anti-weight (use buoyancy or other counter-forces); preliminary anti-action; preliminary action; beforehand cushioning; equipotentiality; reverse action; spheroidality; dynamicity (make rigid parts movable); partial or excessive action; transition to a new dimension; mechanical vibration; periodic action; continuity of useful action; rushing through; "blessing in disguise" (turn harm into benefit); feedback; mediator; self-service; copying; cheap-short-living objects; mechanics substitution; pneumatic and hydraulic constructions; flexible shells and thin films; porous materials; colour change; homogeneity; rejecting/regenerating; transformation of physical and chemical states; phase transitions; thermal expansion; strong oxidants; inert atmosphere; composite materials.

\textbf{Patterns of evolution.} Engineering systems follow recognisable evolutionary patterns: simple to complex; small to large to right-sized; rigid to flexible; rigid to articulated; mono-system to bi-system to poly-system. The patterns can be used to anticipate where a current system can evolve to.

\textbf{Substance-field analysis.} The interactions in a system are modelled as substances acting on each other through fields (mechanical, thermal, chemical, electromagnetic, etc.). The model identifies missing interactions (a desired action that no current substance-field provides) and harmful interactions (an undesired effect that a substance-field produces); the inventive solutions add or modify substances and fields to address them.

TRIZ has been criticised on multiple grounds: the contradiction matrix is empirical and its categories are not always sharp; the inventive principles overlap and can be applied post-hoc to almost any solution; the framework's terminology can mask its substantive insights; the patent base it derives from is biased toward certain industries (manufacturing; mechanical engineering) and certain eras. Despite the criticisms, the framework has substantive value as a generative method: the contradiction-resolution framing is useful even without the full TRIZ toolkit; the inventive principles are a checklist that surfaces options engineers would not generate associatively; the patterns of evolution provide useful design heuristics.

The engineering-first treatment in this chapter uses TRIZ as one generative technique among several; the chapter does not require mastery of the full framework. The reader who finds TRIZ compelling can study Altshuller's "And Suddenly the Inventor Appeared" or the modern TRIZ literature; the reader who does not can use the contradiction-resolution framing without the formal apparatus.

\section{Concept selection: weighted matrix, Pugh chart}\pathtag{core}

Once concepts are generated, the design team selects among them. The selection is not a vote but a structured evaluation against the requirements; the documentation of the evaluation is what makes the selection accountable.

\textbf{Pugh chart.} The Pugh chart (Stuart Pugh's concept-selection method) is the dominant technique. The procedure:

The concepts are listed as columns of a matrix; the requirements (or evaluation criteria) are listed as rows. One concept (often the current solution or a reasonable baseline) is designated the "datum." For each requirement, each non-datum concept is evaluated against the datum: "+" if the concept is better than the datum on this requirement; "0" if equivalent; "$-$" if worse. The columns are summed: the concept with the highest positive total (or the lowest negative total) is the candidate for further development.

The Pugh chart's virtues:

\textbf{Forces explicit comparison.} The matrix structure forces the team to consider every concept against every requirement; the comparison cannot be skipped.

\textbf{Surfaces strengths and weaknesses.} The pattern of +/0/$-$ across requirements shows which concepts win on which dimensions; the patterns suggest hybrid concepts (combine the strengths of two concepts) that no individual concept achieves.

\textbf{Resists numerical false-precision.} The +/0/$-$ scale is intentionally coarse; it avoids the false precision of "concept A scores 7.3 on requirement X" when the true uncertainty is much larger than the difference between 7.3 and 6.8.

\textbf{Documents the rationale.} The completed chart is the record of the selection decision; the rationale is auditable.

The Pugh chart's limitations:

\textbf{All requirements treated equally.} The basic Pugh chart does not weight requirements; a critical requirement is treated the same as a peripheral one. The remediation is to use a weighted matrix (next paragraph) or to apply the Pugh chart with a priority filter (must-have requirements eliminate concepts that fail them, regardless of the count).

\textbf{Datum-dependency.} The selection depends on which concept is the datum; different datums can produce different rankings. The discipline is to repeat the analysis with multiple datums; concepts that win across multiple datums are robustly preferred.

\textbf{Sum-of-pluses-and-minuses is approximate.} The summation treats different requirements as if they were commensurable; they are not. The discipline is to read the pattern of +/0/$-$ rather than to rely on the sum alone.

\textbf{Weighted decision matrix.} The weighted matrix extends the Pugh chart with numerical weights for each requirement. Each concept-requirement cell is scored on a numerical scale (1-5 or 1-10); each row is weighted; the weighted sum is the concept's score. The technique is more elaborate than the Pugh chart and produces a single numerical ranking; the cost is the false precision of the numerical scores and weights.

The discipline of using a weighted matrix includes: choosing weights that reflect substantive priority differences; treating the numerical scores with appropriate uncertainty (the difference between a 7 and a 6 on a 10-point scale may be within the scoring noise); checking that the ranking is robust to plausible weight variations.

\textbf{Hybrid approaches.} Many design teams use the Pugh chart for initial down-selection (eliminating obviously dominated concepts) followed by a more detailed weighted matrix for the finalists. The two-stage approach combines the Pugh chart's resistance to false precision with the weighted matrix's discrimination among close candidates.

\textbf{The selection conversation.} The matrix is a tool for structuring the conversation; the conversation is where the selection actually happens. The team's discussion of why one concept scores differently than another surfaces the substantive issues; the numerical output is the summary of the conversation, not its replacement. Teams that treat the matrix as a black box that produces a decision have surrendered the substantive work to the bookkeeping.

\section{First-pass cost and feasibility estimates}\pathtag{standard}

The conceptual stage requires first-pass estimates of cost and feasibility for each candidate concept. The estimates are not precise (the concept is not detailed enough for precise estimates); they are approximate but informative.

The estimation methods:

\textbf{Analogical estimation.} The estimator finds analogous designs whose cost and feasibility are known; the estimate scales from the analogue. A new vehicle's cost is estimated from previous vehicles of similar specification; the technique works when adequate analogues exist; the discipline is to identify analogies carefully and to adjust for known differences.

\textbf{Parametric estimation.} Cost is modelled as a function of design parameters (mass; complexity; performance; technology level). The parametric models are fit to historical data; new designs are estimated by substituting the design parameters. The technique is widely used in aerospace, automotive, and construction.

\textbf{Bottom-up estimation.} The concept is decomposed into subsystems; each subsystem's cost is estimated; the total is the sum. The technique is more expensive (it requires concept-level detail in each subsystem) but more accurate for designs that depart substantially from prior analogues.

\textbf{Three-point estimation.} Each cost is estimated as best-case, most-likely, worst-case; the expected cost is a weighted average (in PERT, $(O + 4M + P)/6$). The technique acknowledges the uncertainty and produces an estimate with confidence bounds rather than a single number.

\textbf{Feasibility scoring.} For each concept, feasibility along multiple dimensions (technical; manufacturing; regulatory; schedule; supply chain) is scored qualitatively (high/medium/low) or numerically; the scores feed the selection matrix.

The estimates' purpose at the conceptual stage is differential: which concept costs less than which other concept; which is more feasible. The absolute numbers are approximate; the relative comparisons are what informs the selection.

The discipline of first-pass estimation includes: documenting the assumptions; identifying the cost drivers (the dimensions to which the estimate is most sensitive); flagging the uncertainties; updating the estimates as the concept evolves. The estimates are themselves engineering deliverables; they enter the design history and become the basis for subsequent detailed estimates.

\section{Failure: premature commitment to a concept}\pathtag{core}

The dominant failure mode in conceptual design is premature commitment: the team commits to a concept before adequate generation and evaluation; the commitment shapes the detailed design; alternatives that would have been better are not explored.

The mechanisms:

\textbf{First-concept dominance.} The first concept the team generates becomes the working concept; subsequent generation is constrained by the team's investment in the first concept. The discipline of structured generation (Section 1) resists the dominance, but the resistance is not automatic; teams without the discipline default to first-concept commitment.

\textbf{Familiarity preference.} The team prefers the concept similar to what they have done before; the familiar concept is less risky to develop but may be less innovative than alternatives. The preference is rational at the individual project level (familiar concepts have known development paths) but produces aggregate stagnation across the team's portfolio.

\textbf{Sponsor commitment.} A senior team member or project sponsor commits to a specific concept early; the team's subsequent generation is shaped to support the committed concept rather than to surface alternatives. The remediation is structural: explicit separation between generation and selection; multiple-concept presentation requirement for design reviews; sponsor commitment to evaluating alternatives rather than to a specific concept.

\textbf{Sunk-cost commitment.} The team has invested in developing one concept to some level of detail; the investment makes the team reluctant to discard the concept even when subsequent evidence suggests an alternative would be better. The bias is well-documented in human cognition; the discipline of staged commitment (early commitments are tentative; later commitments are firm) is the response.

\textbf{Schedule pressure.} The project schedule does not allow time for generating and evaluating multiple concepts; the team commits to the first plausible concept to make schedule. The dynamic is the production-pressure pattern of Volume X Chapter 10; the cost is the downstream rework when the committed concept proves inadequate.

\textbf{Tool commitment.} The team's design tools (CAD libraries, simulation models, analysis templates) are committed to a particular concept class; alternative concepts that would require different tools are at a disadvantage in evaluation because the team's tool infrastructure does not support them. The remediation is to maintain tool flexibility through the conceptual stage; the commitment to specific tools comes later in detailed design.

The cost of premature commitment is the design space the team failed to explore. The cost is invisible at the project level (the team does not know what they did not generate); it is visible at the portfolio level (the team's designs are conservative and similar) and at the failure level (failures often reveal that the committed concept was inadequate to a known operational condition that an alternative would have addressed).

The structural responses include: stage-gated processes that require evidence of generation breadth before the conceptual stage closes; design-review protocols that require multiple-concept presentation; assignment of devil's-advocate roles to challenge the team's preferred concept; periodic challenge sessions in which an external party reviews the team's conceptual work and surfaces alternatives the team did not generate.

The Boeing 737 MAX MCAS case (Volume X Chapter 8) is partly a premature-commitment failure: the single-AoA architecture was committed early in MCAS's development; subsequent expansion of MCAS authority did not revisit the architectural commitment; the dual-AoA alternative was available technically but was not generated and evaluated at the points in development where the architectural choice could have been re-opened.

\begin{archetype}[optimisation]
Conceptual design is the optimisation archetype's application at the conceptual stage of the design process. The engineer is searching the conceptual design space for the concept that best satisfies the requirements under the constraints. The search is wide (generate many alternatives) and structured (evaluate against the requirements); the substantive engineering judgment is in the generation discipline and the selection criteria.
\end{archetype}

\begin{principle}[Multiple concepts before commitment]
The discipline of conceptual design is the discipline of generating multiple concepts and selecting through structured evaluation before committing to detailed design. Premature commitment surrenders the highest-leverage decision in the design process. The cost of generating ten concepts is small; the cost of the rework when a committed concept proves inadequate is large; the engineering economics favour discipline.
\end{principle}

\begin{project}[Ten concepts, Pugh selection]
Select one design need: a specific user requirement (drawn from the reader's environment, from the projects in earlier chapters, or from a publicly-stated need). Generate ten distinct concepts for the design, each captured in a one-page sketch with annotation explaining the principle of operation. Evaluate the ten concepts against 5-8 requirements using a Pugh chart, with one concept designated the datum. Identify the top-ranked concept and the hybrid possibilities suggested by the pattern of strengths across concepts. Deliverable: 10 sketch pages plus a 2-page selection document with the Pugh chart and rationale.

\textbf{Hazard.} The first few concepts will be obvious; the fifth through tenth are the substantive contribution. Stopping at three concepts because "we already have a good one" is the discipline-failure this project is designed to address.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A morphological matrix has 7 functional dimensions with 4, 5, 3, 6, 4, 5, 3 options respectively. Compute the total concept space size. Compute the size after feasibility filtering eliminates 80\% of combinations. Discuss the implications for evaluation effort.
\end{exercise}

\begin{exercise}
A weighted decision matrix evaluates 5 concepts against 8 requirements with weights summing to 1. Each concept-requirement score is in $[0, 10]$. Compute the maximum possible weighted score and the score for a concept that scores 7 on every requirement.
\end{exercise}

\begin{exercise}
The PERT three-point estimation formula is $(O + 4M + P)/6$. For a concept estimated at $O = \$10M$, $M = \$15M$, $P = \$25M$, compute the expected cost. Compute the standard deviation (PERT uses $(P-O)/6$). Discuss the implication for concept comparison when two concepts have overlapping cost distributions.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
For a Pugh chart with $N$ concepts and $M$ requirements, derive the conditions under which the ranking is robust to the choice of datum. Identify the structural properties of the concept-requirement matrix that produce robust rankings.
\end{exercise}

\begin{exercise}
Derive the expected number of distinct concepts a brainstorming team of $K$ members generates, given each member generates $n$ ideas and there is overlap probability $p$ between any two ideas from different members. Identify the regime in which adding members improves generation versus the regime in which redundancy dominates.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the number of distinct concepts a typical product-development team explicitly considers for a new product before committing to one. Identify the implication of the gap between the actual number and the discipline's recommendation of ten or more.
\end{exercise}

\begin{exercise}
Estimate the cost of generating an additional five concepts beyond a team's natural stopping point (e.g., going from 3 to 8 concepts). Compare to the cost of the rework when a committed concept proves inadequate in detailed design.
\end{exercise}

\begin{exercise}
Estimate the fraction of major engineering products that go to market with a concept that the development team substantially modified or replaced after the initial concept commitment. Discuss what the fraction implies about the discipline of conceptual design as actually practised.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Implement a simulation of concept generation under the first-concept-dominance bias: each generated concept has probability $\beta$ of being constrained to resemble the first concept. Sweep $\beta$ and identify the conditions under which generation produces effective diversity versus the conditions under which the bias dominates.
\end{exercise}

\begin{exercise}
Simulate a Pugh chart selection in which the underlying "true" concept quality has known values but the team's evaluation contains noise. Sweep the noise level and identify the conditions under which the selection identifies the best concept with high probability.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a conceptual-design review protocol for an engineering organisation. Specify what the team must present, what the reviewers evaluate, what the criteria for passing are, and what the consequences of a failed review are.
\end{exercise}

\begin{exercise}
Design a brainstorming-session template for a design team. Specify the facilitation approach, the time allocation, the documentation, and the integration of session output with the subsequent down-selection process.
\end{exercise}

\begin{exercise}
Design a morphological-analysis worksheet for a specific design problem (e.g., a personal urban transport device). Specify the functional dimensions, the candidate options per dimension, and the feasibility filtering criteria.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: the discipline of generating ten concepts before committing is misallocated effort for routine design problems where the natural solution is well-established. Argue the opposite. Identify what distinguishes the two regimes and how the engineer should judge which regime applies.
\end{exercise}

\begin{exercise}
Discuss the role of senior-engineer intuition in concept selection. Argue: experienced engineers' rapid concept selection is reliable and resists premature-commitment criticism. Argue the opposite: experience produces systematic blind spots that structured generation addresses. Identify the empirical evidence that bears on the dispute.
\end{exercise}

\begin{exercise}
TRIZ has been criticised as post-hoc rationalisation of solutions that engineers would have generated without the framework. Discuss the criticism. Identify the conditions under which TRIZ produces substantive value versus the conditions under which it produces theatre.
\end{exercise}

\begin{exercise}
The 737 MAX MCAS architecture was committed early and was not re-evaluated when the system's authority expanded \cite{acc:house-737max-2020}. Discuss what conceptual-design discipline would have caught the inadequacy and what institutional structures would have made the discipline operate as designed.
\end{exercise}

\begin{exercise}
A design organisation's project portfolio shows that 80\% of their products use the same concept family. Discuss what the pattern implies about the organisation's conceptual-design practice and what changes would broaden the conceptual space the organisation explores.
\end{exercise}

\begin{exercise}
The Pugh chart's selection ranking can change when the datum concept changes. Discuss whether this property is a strength or a weakness of the technique. Identify a structured way to use the property to surface concepts whose value depends on the comparison baseline (rather than as a defect to be eliminated).
\end{exercise}

\begin{exercise}
A senior engineer presents a concept at the project kickoff and says "this is the obvious solution; let's just go with it." Discuss the engineer's options for responding in a way that preserves the conceptual-design discipline without antagonising the senior colleague. Identify the institutional structures that would protect the discipline even when the senior engineer's preference is strong.
\end{exercise}
