\chapter{Coupled human-technical systems}
\label{vol09:ch16}

\epigraph{The system is the operator plus the machine; the engineer who designs only the machine has designed half the system.}{}

\chapmeta{Half-life: durable. Archetypes: interface, coupling, failure. Exercise target: 25. Project track: physical.}

\noindent
Engineered systems are operated by humans: pilots fly aircraft; surgeons use surgical robots; operators monitor industrial control rooms; drivers interact with semi-autonomous vehicles; doctors interpret diagnostic equipment; users configure consumer products. The interaction between the human and the technical system is itself part of the engineering: the operator's mental model, attention, training, fatigue, and motivation all shape the system's behaviour. Failures of human-technical coupling are responsible for a substantial fraction of accidents in aviation, medicine, process industries, and emerging autonomous systems. This chapter develops the operator in the loop, human factors and ergonomics, Bainbridge's automation paradox, levels of automation, mode confusion and shared control, trust calibration, and the major case studies of human-technical coupling failures.

\section{The operator in the loop}\pathtag{core}

The operator is the human who interacts with the engineered system. The interaction takes many forms: monitoring (the operator observes the system's state), supervising (the operator authorises or overrides automatic actions), controlling (the operator commands the system directly), maintaining (the operator services the system), troubleshooting (the operator diagnoses and responds to abnormalities), interpreting (the operator extracts meaning from system outputs).

The operator's role is shaped by the system design: a fully manual system requires the operator to perform all control actions; a fully automatic system reduces the operator to monitoring; intermediate levels share control between operator and automation. The choice of operator role is an engineering decision; the choice affects performance, training, attention, fatigue, and failure modes.

The operator brings capabilities that machines lack: judgement in novel situations, pattern recognition across unstructured contexts, ethical and social reasoning, accountability. The operator also brings limitations: bounded attention, cognitive biases, fatigue, training variability, emotional state. The engineering challenge is leveraging the operator's strengths while accommodating their limitations.

The operator is part of the system, not external to it. The system's reliability is jointly determined by the technical components and the operator; the operator's reliability under given conditions is part of the system's reliability. This realisation shifts the engineering question from "design the machine" to "design the combined human-machine system."

\begin{archetype}[interface: human-machine coupling as engineering surface]
The human-machine interface is the engineering surface where the operator's capabilities meet the system's capabilities. The interface design determines what the operator can perceive, what they can decide, what they can command, what they can learn from. A well-designed interface presents the right information at the right time in the right form; a poorly-designed interface produces confusion, errors, and failures. Interface design is engineering; the discipline includes human factors, ergonomics, cognitive psychology, and operational practice.
\end{archetype}

\section{Human factors and ergonomics}\pathtag{standard}

Human factors engineering applies knowledge of human capabilities and limitations to the design of systems. The discipline includes physical ergonomics (posture, reach, anthropometry), cognitive ergonomics (perception, attention, memory, decision-making), and organisational factors (team dynamics, culture, training).

Standard concerns:

\textbf{Display design:} how information is presented to the operator. Standard guidelines: relevant information emphasised; clutter minimised; layout consistent; coding (colour, shape, position) systematic; alarms differentiated by priority. Modern aviation displays follow these principles; many industrial control rooms do not.

\textbf{Control design:} how the operator acts on the system. Standard guidelines: control-display compatibility (the control's direction matches the display's direction); feedback (the operator perceives the control's effect); affordances (the control's shape suggests its function); guards (critical controls protected from accidental actuation).

\textbf{Workload:} the demands placed on the operator's perceptual, cognitive, and motor capacities. Excessive workload produces errors; insufficient workload produces vigilance loss. The Yerkes-Dodson law (1908): performance is maximised at intermediate workload.

\textbf{Situational awareness:} the operator's perception, comprehension, and projection of the situation. Endsley's (1995) three-level model: level 1 (perception of the elements); level 2 (comprehension of the situation); level 3 (projection of future status). Loss of situational awareness is a contributing factor in many incidents.

\textbf{Workload measurement:} subjective (NASA-TLX, modified Cooper-Harper), physiological (heart-rate variability, pupillometry, eye-tracking), performance-based (task time, error rate, secondary task). Each measure has strengths and weaknesses; combination is often most informative.

\textbf{Training:} the operator's capability is acquired through training. The discipline includes task analysis (what the operator must do), instructional design (how training is delivered), evaluation (whether training is effective), and currency (how training is maintained).

Standards include MIL-STD-1472 (military), HFES (Human Factors and Ergonomics Society) standards, ANSI/HFES standards, IEC 62366 (medical devices), ISO 26800 (general). The standards encode accumulated practice; new applications require adaptation, but the foundational principles transfer.

\section{Automation surprises (Bainbridge's paradox)}\pathtag{standard}

Lisanne Bainbridge's 1983 paper "Ironies of Automation" identified a paradox: the more automation an operator is given, the more the operator's role becomes monitoring and exception-handling; but human capability for monitoring decays without practice, and exception-handling requires the operator to step in to handle precisely the conditions automation cannot. The operator is trained to do nothing; when they must act, they are out of practice and the conditions are difficult.

The paradox extends. Automated systems are often designed for the easy cases; the operator handles the edge cases. The operator's training emphasises normal operation; the actual operator engagement is the abnormal. The system's performance metrics measure automation efficiency; the operator's contribution is invisible until needed.

Modern responses to the paradox:

\textbf{Adaptive automation:} the system's automation level varies based on the operator's workload, fatigue, or capability. The operator stays engaged at intermediate levels; full automation only when monitoring is reliable.

\textbf{Operator-in-the-loop training:} the operator practices the hand-flying skills, the manual procedures, the troubleshooting. Currency requirements mandate periodic hands-on practice. Aviation has these traditions; industrial automation often does not.

\textbf{Automation transparency:} the operator can see what the automation is doing and why. Modern displays show the automation's state, mode, intentions, and reasoning; the operator can anticipate and intervene.

\textbf{Graceful degradation:} when automation fails, it does so in a way that preserves operator capability. A clear handover, with notification of the new responsibilities and information about the failure, is essential.

\textbf{Joint cognitive systems:} the engineering frames operator and automation as a joint system, optimised together. The operator's tasks and automation's tasks are designed jointly, not by automating what is easy and leaving the rest to the operator.

\section{Levels of automation}\pathtag{standard}

The level of automation describes how much the system does versus the operator. Standard taxonomies:

\textbf{Sheridan-Verplank (1978):} 10 levels from "computer offers no assistance" to "computer decides everything and acts autonomously."

\textbf{Parasuraman-Sheridan-Wickens (2000):} four-stage model (information acquisition, information analysis, decision selection, action implementation), each with multiple levels.

\textbf{SAE J3016 (driving automation):} levels 0-5, from no automation to full driving automation. The taxonomy is widely cited in autonomous-vehicle discussion.

The choice of automation level is an engineering decision with significant consequences. Level $0$ (no automation) places full burden on the operator. Level $10$ (full automation) eliminates operator engagement. Intermediate levels share control but introduce mode-confusion risks.

The 1979 Three Mile Island accident (\cite{acc:kemeny-tmi-1979}) involved operators acting on incomplete and misleading information from automation; the indicators showed the position command, not the actual valve position; the operators inferred (incorrectly) that the valve was closed for several hours. The accident's investigation recommended significant changes to control-room design and operator training.

The 2018-2019 Boeing 737 MAX accidents involved an automation system (MCAS, the Maneuvering Characteristics Augmentation System) that pushed the nose down based on angle-of-attack sensor input; the pilots, not adequately trained on the system's behaviour, were unable to recover from runaway MCAS activations triggered by sensor failure. The accidents grounded the 737 MAX fleet for $20$ months and prompted significant regulatory and design changes.

\section{Mode confusion and shared control}\pathtag{core}

A system mode is a state in which the system behaves differently from other states. Examples: an autopilot in "altitude hold" mode behaves differently from "vertical speed" mode; a car in "cruise control on" vs "off"; a digital therapy machine in "X-ray" vs "electron beam" mode; an aircraft in "auto-throttle" vs "manual throttle."

Mode confusion is the operator's misunderstanding of which mode the system is in, leading to actions appropriate to the wrong mode. The consequences range from minor inconvenience to catastrophic failure. Classic examples:

\textbf{Therac-25:} the operator selected electron-beam mode then quickly changed to X-ray mode; the software took a different code path; the patient received massive radiation overdose. The mode confusion was the software's, but the operator's interface did not reveal the inconsistent state.

\textbf{A320 Mont Sainte-Odile (1992):} the crew set vertical descent rate to $-3300\,\text{ft/min}$ when intending $-3.3^\circ$ descent angle; the autopilot mode interpreted the input one way, the crew the other. The aircraft descended into terrain.

\textbf{Air France 447 (\cite{acc:bea-af447}, 2009):} the flight envelope protection disengaged due to inconsistent pitot data; the crew's response was suited to the protected mode rather than the manual mode they had entered. The aircraft stalled and was lost.

\textbf{Tesla Autopilot (multiple events):} drivers misunderstand the system's capability boundaries; engaging Autopilot in conditions where it cannot operate safely; failing to maintain attention; resulting accidents include collisions with parked vehicles, emergency vehicles, and concrete barriers. The events span multiple years and continue.

Shared control distributes commands between operator and automation. The standard architectures:

\textbf{Trading control:} one party (operator or automation) has full control at any time; control is transferred at defined transitions.

\textbf{Cooperative control:} operator and automation control different aspects simultaneously. Lane-keeping (automation) plus speed (operator) is a familiar example.

\textbf{Negotiated control:} the operator and automation negotiate the control authority dynamically. More research-grade than deployed.

The engineering of shared control requires explicit consideration of the mode structure: when is the operator in control, when is the automation, what triggers transitions, what does the operator need to know at each transition, how is the operator brought back into the loop.

\section{Trust calibration in automated systems}\pathtag{standard}

The operator's trust in the automation determines their reliance on it. Trust should be calibrated: high enough to use the automation when it works, low enough to override when it does not. Over-trust (using automation when it fails) and under-trust (overriding automation that works) both produce poorer performance than calibrated trust.

Factors that shape operator trust:

\textbf{Performance experience:} the operator's history with the system. Reliable performance builds trust; failures reduce it. The shape of the trust curve over experience is complex.

\textbf{Transparency:} the operator can see how the automation works and why it acts. Transparent automation supports calibrated trust; opaque automation produces over-trust (when it works) or under-trust (when it fails surprisingly).

\textbf{Feedback:} the operator can observe the consequences of the automation's actions. Continuous feedback supports calibration; missing feedback produces drift.

\textbf{Training:} the operator's understanding of the automation's capabilities and limitations. Inadequate training produces miscalibrated trust.

\textbf{Operator state:} fatigue, distraction, cognitive load all shift trust calibration. The trust calibration that an alert operator achieves may not survive the night shift.

Engineering responses to trust calibration include: explicit display of automation confidence; explicit identification of automation limits; periodic operator-in-the-loop training; design that resists complacency (occasional engagement requirements). The challenge is that miscalibrated trust does not manifest until a failure occurs; the engineering response is preventive.

\begin{estimation}
\textbf{Question.} Estimate the operator engagement that is needed to maintain a useful level of attention for an automated system that handles routine operations reliably.

\textbf{Working.} The vigilance literature suggests that human attention to a passive monitoring task degrades significantly after $20$ to $30$ minutes; the operator's performance on a detection task is much worse after an hour than at the start. Engineering responses: rotate the operator's role periodically (the system's automation level varies with time); inject test events that require operator response; require periodic operator action even when not strictly needed; pair operators so one is always more engaged. The level of engagement required to maintain calibration is approximately one meaningful interaction per $10$ to $20$ minutes; less and the operator is effectively absent from the loop, with the consequences discussed above.
\end{estimation}

\section{Failure: Three Mile Island, Air France 447, Tesla Autopilot incidents}\pathtag{core}

Human-technical coupling failures have characteristic patterns.

\textbf{Three Mile Island (1979):} the partial meltdown of Reactor 2 at the Three Mile Island Nuclear Generating Station, USA. The accident sequence began with a stuck-open pressure-operated relief valve; the operators inferred (from misleading indicators) that the valve was closed; they overrode high-pressure injection that would have stabilised the reactor; the resulting partial core melt was contained but the political consequences were profound. The Kemeny Commission report (\cite{acc:kemeny-tmi-1979}) identified deficiencies in operator training, control-room design, plant procedures, and regulatory oversight. The accident reshaped commercial nuclear power.

\textbf{Air France 447 (2009):} the loss of an Air France Airbus A330 over the Atlantic, with $228$ fatalities. The aircraft entered an area of severe weather with ice-blocked pitot tubes; the airspeed indications became unreliable; the autopilot and flight envelope protection disengaged; the pilots, operating outside their normal envelope and with confusing displays, applied inappropriate inputs that triggered a sustained high-altitude stall. The investigation (\cite{acc:bea-af447}) identified failures of cockpit indicators, pilot training, and crew resource management.

\textbf{Tesla Autopilot (multiple events):} as discussed in the section on mode confusion. The driver-monitoring system has been adjusted over years; the public discussion of "Autopilot" and "Full Self-Driving" naming continues; the regulatory and legal responses are still being shaped.

\textbf{The 2017 Frankel et al. autonomous-vehicle accident review (multiple events):} the Uber pedestrian fatality, the Tesla truck-crossing accidents, the various consumer-vehicle incidents share the pattern: the system's capability boundary was less than the operator's understanding; the operator was not adequately engaged to intervene; the failure occurred at the boundary.

\textbf{2018 Lion Air 737 MAX and 2019 Ethiopian Airlines 737 MAX:} the MCAS-driven crashes that led to the worldwide 737 MAX grounding for $20$ months. The pilots had not been adequately trained on MCAS; the simulator training was inadequate; the response procedure was difficult to execute; the AoA-sensor failure on Lion Air and the wiring-related sensor failure on Ethiopian Airlines were not on the published Boeing list of failure modes. The accidents led to significant Boeing reorganisation, FAA reform, and ongoing certification review of the MAX and other types.

\textbf{Common patterns:} the operator did not know what the automation was doing; the automation did not know what the operator was thinking; the system as a whole behaved in a way that neither half intended; the failure was the boundary, not any single component.

\textbf{Engineering responses:}

\textbf{Transparency:} the operator can see what the automation is doing.

\textbf{Predictability:} the automation behaves consistently with the operator's mental model.

\textbf{Graceful degradation:} the transitions between modes are clear and well-trained.

\textbf{Training:} the operator has practiced the abnormal as well as the normal.

\textbf{Continuous monitoring:} the operator's engagement is observable and required.

\textbf{Limits acknowledged:} the automation's capability boundary is part of the operator's training and the operating procedure.

\textbf{Investigation discipline:} the failure modes are catalogued and the responses are organisationally learned.

\begin{principle}[The joint-cognitive-system principle]
The operator and the automation are a joint cognitive system; the engineering must address them together. The operator's perception, attention, decision-making, and action are part of the system, not external to it. The automation's transparency, predictability, and limits are part of the operator's perception, not separate from it. The engineering discipline is to design the joint system: the operator's task and the automation's task allocated jointly; the operator's interface tailored to the operator's capabilities; the automation's behaviour designed for the operator's understanding; the failure modes anticipated and trained. The half-system approach (design the machine and leave the operator to figure it out) is the source of many accidents.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Redesign the interface of one piece of household automation]
\textbf{Objective.} Redesign the user interface of one piece of household automation (thermostat, washing machine, smart speaker, smart lighting, security system, EV charger). Defend the design choices in terms of human-factors principles. Compare to the existing interface.

\textbf{Method.} Choose a piece of automation. Document the current interface: physical controls, displays, modes, feedback, alarms. Identify the human-factors problems: mode confusion, opaque automation, poor feedback, excessive workload, hidden state. Redesign the interface: layout, controls, displays, automation behaviour, mode indicators, feedback. Make the redesign concrete: drawings, mockups, written explanation. Compare the redesigned interface to the original.

\textbf{Deliverables.} The original interface description and identified issues; the redesigned interface (drawings or mockups); the human-factors rationale for each design choice; a comparison highlighting the expected improvements; a written discussion of trade-offs.

\textbf{Hazard.} The project is paper-design and analysis; no implementation hazards.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A control room has $200$ alarms; the operator can handle approximately $10$ alarms per hour. Compute the maximum manageable alarm rate during normal operation.
\end{exercise}

\begin{exercise}
A vigilance task has $1\%$ event rate; after $1$ hour, detection probability is $80\%$. Compute the expected number of missed events per hour.
\end{exercise}

\begin{exercise}
A pilot's reaction time is $2\,\text{s}$; the aircraft descends at $1000\,\text{ft/min}$. Compute the altitude lost during the reaction.
\end{exercise}

\begin{exercise}
An autonomous-vehicle's perception system has $50\,\text{ms}$ latency, the controller has $20\,\text{ms}$ processing time, and the brakes have $200\,\text{ms}$ actuation. At $60\,\text{km/hr}$, compute the stopping distance after detection.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the trust-calibration condition in terms of operator detection probability, false-alarm rate, and reliance behaviour.
\end{exercise}

\begin{exercise}
Derive the operator-workload relationship between detection performance and event rate for a vigilance task.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the operator engagement required to maintain situational awareness in a supervisory-control task with $5\%$ off-nominal events.
\end{exercise}

\begin{exercise}
Estimate the training time required for an operator to develop reliable manual-control skills in a previously-automated task.
\end{exercise}

\begin{exercise}
Estimate the alarm-management threshold below which alarm flooding becomes operationally disabling.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate operator vigilance in a monitoring task with stochastic events; characterise the detection performance over time.
\end{exercise}

\begin{exercise}
Simulate an automation transition (e.g., autopilot disengagement) under different operator-readiness conditions; compute the consequence of the readiness state.
\end{exercise}

\begin{exercise}
Simulate a trust-calibration scenario: an operator interacts with an automation of varying reliability; observe the resulting reliance patterns.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design the cockpit alert hierarchy for a commercial aircraft: cautions, warnings, master cautions, master warnings, recovery alerts.
\end{exercise}

\begin{exercise}
Design a driver-monitoring system for a Level 2 autonomous-driving feature: monitoring methods, response thresholds, fallback behaviour.
\end{exercise}

\begin{exercise}
Design a control-room information display for a chemical-process control room: information hierarchy, alarm management, situational displays.
\end{exercise}

\begin{exercise}
Design a training and competency framework for nuclear-plant control-room operators.
\end{exercise}

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A team's autopilot has been disengaging in marginal weather; pilots report confusion. State the engineering response.
\end{exercise}

\begin{exercise}
A team's industrial control room is experiencing alarm flooding during upset conditions. State the engineering response.
\end{exercise}

\begin{exercise}
A team's product team wants to expand the operating envelope of a semi-autonomous feature. State the engineering considerations for operator training, capability limits, and trust calibration.
\end{exercise}

\begin{exercise}
A team is choosing the level of automation for a new medical-device feature. State the engineering criteria.
\end{exercise}

\begin{exercise}
A team's pilot training emphasises automation use; recent incidents suggest pilots are not maintaining manual skills. State the engineering and training response.
\end{exercise}

\begin{exercise}
A team's autonomous-vehicle perception system performs reliably; the safety analysis shows the operator is unlikely to be ready when an intervention is needed. State the engineering judgement and design options.
\end{exercise}

\begin{exercise}
A team's HMI shows the automation's state but not its intentions; operators have reported being surprised by automation actions. State the engineering response.
\end{exercise}

\begin{exercise}
A team's training-effectiveness review shows that infrequently-trained scenarios (emergency response) are inadequately retained over time. State the engineering response.
\end{exercise}

\begin{exercise}
A team is asked to certify a Level 3 autonomous driving system. State the engineering considerations including the human-factor analysis of the operator handover during automation disengagement, the role of driver monitoring, and the relevant certification standards.
\end{exercise}
