\chapter{Calibration and traceability}
\label{vol01:ch03}

\epigraph{A measurement result earns public meaning when its value
and uncertainty can be related to a specified reference.}{A
metrological rule in the vocabulary of the VIM
\cite{std:jcgm-vim} and the NIST traceability policy
\cite{std:nist-traceability}.}

\chapmeta{Half-life: foundational. Archetypes: interface (the
calibration interface, the chapter's first explicit interface
archetype in Volume I). Exercise target: 30. Project track: analysis
with optional household component. Hazard class: none. Reader path
default: standard, with sections explicitly tagged. Named cases:
Hubble Space Telescope primary mirror (1990), a verified named
accident in the registry; international prototype of the kilogram
(1889--2019), a standards-history case rather than an accident.}

\noindent
A measurement reports a number with a unit. The number's meaning
comes from a chain of comparisons that connects the reader's
instrument to a definition fixed in international agreement. This
chapter describes that chain: how a calibration is performed, how
working laboratory instruments inherit their meaning from primary
standards, how primary standards inherit theirs from defining
constants, and how the entire structure stays trustworthy across
decades, jurisdictions, and orders of magnitude.

\section{What a calibration is}\pathtag{core}

\engterm{Calibration} is the operation that establishes a relation
between an instrument's indication and quantity values provided by
measurement standards, with uncertainties stated on both sides of
the relation. The international vocabulary of metrology defines
calibration as the operation that, under specified conditions, in a
first step establishes a relation between the quantity values with
measurement uncertainties provided by measurement standards and
corresponding indications with associated measurement uncertainties,
and in a second step uses this information to establish a relation
for obtaining a measurement result from an indication
\cite{std:jcgm-vim}.

The definition has two steps.

A calibration begins as a comparison. The instrument produces an
indication; the reference provides a value with stated uncertainty;
the calibration records the relation between the two. Adjustment is
a separate operation. It changes the instrument after calibration
has shown what change is needed.

The report should identify the instrument, the reference, the date,
the conditions, the procedure, the indications, the corresponding
reference values, and the uncertainties. A recalibration interval
may appear on the certificate when required by the customer,
regulator, or quality system. Without records of calibration and
use, the owner has an instrument, but not enough evidence to claim
traceability for a measurement result.

A calibration also depends on the reference's documented status. The
laboratory voltage source is calibrated against a transfer standard;
the transfer standard is calibrated through a higher-level electrical
standard; the primary realisation is tied to Josephson and quantum
Hall effects and the SI constants. Each undocumented link adds an
undocumented uncertainty to the result below it.

\begin{archetype}[Interface]
The interface archetype appears here as the calibration interface.
An interface is a place where a physical quantity becomes a
representation that another system can use. A calibration interface
relates a mass, voltage, length, frequency, or temperature to an
instrument indication, with uncertainty attached to the relation. If
the interface is omitted, the account has not explained how the
quantity became usable.
\end{archetype}

\subsection{Calibration is not adjustment, and is not verification}

Three operations adjacent to calibration are routinely confused with
it. We distinguish them by their purpose:

\begin{itemize}[leftmargin=*]
  \item \engterm{Calibration} establishes a relation between
    indication and reference. It produces information about the
    instrument; it does not change the instrument.
  \item \engterm{Adjustment} changes the instrument so that its
    indication corresponds, under specified conditions, to the value
    of the quantity being measured. Adjustment changes the
    instrument; it requires a prior calibration to know what change
    to make.
  \item \engterm{Verification} provides objective evidence that the
    instrument fulfils specified requirements. Verification
    compares calibration results against specified tolerances; it
    decides whether the instrument is acceptable for use.
\end{itemize}

A calibration produces a relation; an adjustment changes a setting;
a verification produces a yes-or-no judgement against a
specification; validation asks whether the measurement is fit for
its intended use.

\input{volumes/01-quantity/ch03-calibration-and-traceability/figures/fig-calibration-vs-adjustment.tex}

\section{Primary, secondary, working standards}\pathtag{core}

A \engterm{measurement standard} is a realisation of the definition
of a quantity, with a stated value and measurement uncertainty, used
as a reference. Some standards are artifacts, some are instruments
or systems, and some are realised through quantum or thermodynamic
effects. Standards form a hierarchy.

\input{volumes/01-quantity/ch03-calibration-and-traceability/figures/fig-traceability-pyramid.tex}

\subsection{Primary standards}

A \engterm{primary standard} is a standard whose value has the
highest metrological qualities and whose value is accepted without
reference to other standards of the same quantity. In the post-2019
SI, primary standards realise the seven defining constants. The
realisation hardware is operated by national metrology institutes
(NMIs) and designated institutes, with mutual comparisons among them
\cite{std:bipm-si2019, std:bipm-cipm-mra}.

Examples of primary realisations under the post-2019 SI:

\begin{itemize}[leftmargin=*]
  \item The kilogram, realised through Kibble balances (which compare
    mechanical and electrical power) and through X-ray-crystal-
    density measurements of $^{28}$Si single crystals. Either method
    expresses mass in terms of the Planck constant
    $h$ \cite{std:bipm-si2019, web:nist-kibble-balance}.
  \item The metre, realised through frequency-stabilised laser
    interferometry against the cesium-133 hyperfine transition; the
    metre's definition is via the speed of light $c$ and the second
    \cite{std:bipm-si2019}.
  \item The second, realised through cesium-133 fountain clocks
    operating at major NMIs (NIST, NPL, PTB, and others); cesium
    fountain primary standards report fractional uncertainties on
    the order of $10^{-16}$ \cite{web:nist-cesium-fountain}.
  \item The volt, realised through Josephson voltage standards: arrays
    of Josephson junctions whose voltage steps depend only on the
    elementary charge and the Planck constant.
  \item The ohm, realised through the quantum Hall effect.
  \item The kelvin, realised through several primary thermometry
    methods (acoustic gas thermometry, dielectric-constant gas
    thermometry, radiation thermometry), each tied to the Boltzmann
    constant \cite{std:bipm-si2019, web:bipm-cct-its90}.
\end{itemize}

Primary realisation is specialised work. National metrology
institutes and designated institutes maintain selected realisations,
compare them internationally, and publish calibration and measurement
capabilities. The dissemination system rests on those realisations,
their comparisons, and the documented uncertainty budgets that
connect them to working laboratories \cite{std:bipm-cipm-mra}.

\subsection{Secondary standards}

A \engterm{secondary standard} is a standard whose value is
established by calibration with respect to a primary standard or
another higher-order reference for the same quantity. It
disseminates the unit through a documented calibration chain rather
than by definitional authority. Secondary standards may be physical
artifacts (a mass set, a voltage source, a length gauge, a frequency
reference) or laboratory instruments whose values are established by
calibration against a higher reference.

A secondary standard's calibration is recorded on a certificate
that names the primary standard, the date and conditions of
calibration, the measured value, the uncertainty, and the recommended
recalibration interval. A secondary standard with a current,
NMI-issued calibration certificate is the ordinary means by which a
working laboratory acquires access to the SI.

\subsection{Working standards}

A \engterm{working standard} is a standard used routinely to
calibrate or check the working instruments in a laboratory. Working
standards are calibrated against secondary standards (or, in some
laboratories, directly against primary standards), and they in turn
calibrate the laboratory's working instruments.

Most calibration laboratories operate at the working-standard layer.
A calibration laboratory in a manufacturing plant calibrates the
plant's instruments against its working standards; the working
standards are calibrated against secondary standards held by an
external accredited calibration body; the secondary standards are
calibrated against primary standards at an NMI; the primary
standards realise the SI defining constants.

A simple teaching chain has three to five visible links, but real
chains vary. A user may have a short chain through a direct NMI
calibration, or a longer chain through accredited laboratories,
reference standards, transfer standards, and internal checks. A
reader who has done a calibration in a working laboratory has
touched only the working-standard end. A reader who has done a
calibration at an accredited body has touched the secondary end. A
reader who has worked at an NMI has touched the primary end.

\subsection{Historical sketch: how the institutions came to exist}

The standards hierarchy described above is a recent construction.
For most of human technological history, weights and measures were
local: each town, each guild, each market kept its own copies of a
pound, a yard, an inch, a bushel. Trade across boundaries required
interpretation, dispute resolution, and frequent fraud. The modern
international metrology system rests on three institutional
inventions of the late nineteenth and twentieth centuries, plus the
quantum-physics-based redefinition of 2019.

The first invention was the Metre Convention, signed in Paris on
20 May 1875 by seventeen states
\cite{hist:ch03t3-metre-convention-1875}. The treaty established
three bodies: the General Conference on Weights and Measures
(CGPM), a political body of member states that adopts structural
decisions; the International Committee for Weights and Measures
(CIPM), a smaller scientific committee; and the Bureau
International des Poids et Mesures (BIPM), the international
laboratory at the Pavillon de Breteuil in S\`evres
\cite{web:ch03t3-bipm-about}. The Metre Convention's original
charge was a single artifact-based metre and kilogram, but the
institutional architecture proved general enough to extend across
seven base quantities and to survive the 2019 redefinition.

The second invention was the national metrology institute. The
United States created the National Bureau of Standards by Act of
Congress on 3 March 1901 \cite{hist:ch03t3-nbs-act-1901}, signed
by President William McKinley. (The common misattribution to
Theodore Roosevelt reflects Roosevelt's later identification with
Progressive-Era institution-building; Roosevelt did not assume the
presidency until September 1901.) NBS was renamed the National
Institute of Standards and Technology (NIST) in 1988
\cite{web:ch03t3-nist-history-overview}. The United Kingdom had
established its National Physical Laboratory (NPL) in 1900;
Germany's Physikalisch-Technische Reichsanstalt (PTR, founded
1887, now PTB) preceded both. By the 1920s every industrial state
operated a metrology institute, with mutual recognition of
calibration certificates among them.

The third invention was post-WWII institutional rebuilding. The 9th
CGPM in 1948 began the work of cleaning up the proliferation of
unit systems that the war had left behind
\cite{hist:ch03t3-cgpm-9th-1948}. The 11th CGPM (1960) adopted the
Syst\`eme international d'unit\'es as the universally-recognised
system of seven base units \cite{hist:ch03t3-cgpm-11th-1960}; the
metre was defined that year as the wavelength of a krypton-86
discharge transition (superseded in 1983 by the speed-of-light
definition). The 13th CGPM (1967) replaced the ephemeris-second
definition with the atomic-second: the cesium-133 hyperfine
transition at 9\,192\,631\,770 Hz exact
\cite{hist:ch03t3-cgpm-13th-1967, text:ch03t3-mccoubrey-1991}. The
14th CGPM (1971) added the mole as the seventh base unit
\cite{hist:ch03t3-cgpm-14th-1971-mole}. The 17th CGPM (1983)
redefined the metre via the speed of light
\cite{hist:ch03t3-cgpm-17th-1983-metre}. Each redefinition pinned
a previously-measurable quantity to a previously-measured constant,
shrinking the set of independent measurements required to disseminate
the SI.

The fourth invention, the 2019 redefinition, completed the
programme. The 26th CGPM in 2018 fixed the numerical values of the
Planck constant, the elementary charge, the Boltzmann constant, and
the Avogadro number; the kilogram, ampere, kelvin, and mole became
realisable from those constants alone \cite{std:bipm-si2019}. The
international prototype of the kilogram, the last surviving artifact
definition in the SI, was demoted from defining standard to working
reference. The dissemination chain described in this chapter is the
operational realisation of that history. A reader's calibration
certificate from NIST in 2026 traces its authority through a chain
that runs back, institutionally, to a treaty signed in 1875 and,
technically, to four constants of physics fixed in 2018.

The post-WWII metrology infrastructure was the precondition for
global manufacturing as a feasible practice. A bolt thread produced
in one country fits a nut produced in another because both were
manufactured against gauges traceable to NMI-issued length
standards; an electronic component drawing a current specified by
its datasheet matches measurement in a laboratory on the other side
of the world because both currents are realised through the same
Josephson-and-quantum-Hall basis. The MRA (1999), accredited
laboratory networks under ISO/IEC 17025, and the CIPM key-comparison
programme are the institutional machinery that maintains the
correspondence \cite{std:bipm-cipm-mra, std:iso-iec-17025}.

\section{The traceability chain and NIST (and BIPM)}\pathtag{core}

\engterm{Metrological traceability} is the property of a measurement
result whereby the result can be related to a reference through a
documented unbroken chain of calibrations, each contributing to the
measurement uncertainty \cite{std:jcgm-vim, std:nist-traceability}.
The chain is the operational realisation of the standards hierarchy
described above.

A traceable measurement result is one whose calibration certificate
can be followed link-by-link from the working instrument back to a
primary realisation. Each link in the chain has its own calibration
certificate. Each certificate names the higher-level reference and
contributes its own uncertainty to the cumulative uncertainty of the
final measurement. The chain's existence is what distinguishes a
measurement from an opinion; the chain's documentation is what
makes the measurement defendable in a courtroom, in a regulatory
audit, or in a dispute between two engineers.

\subsection{Who maintains the chain}

The chain is maintained by a global system of institutions, in
descending order of authority:

\begin{itemize}[leftmargin=*]
  \item The Bureau International des Poids et Mesures (BIPM), in
    S\`evres, France. The BIPM is the international body
    coordinating metrology under the Metre Convention of 1875. It
    maintains some primary realisations directly, conducts
    international comparisons among national metrology institutes,
    and operates the CIPM Mutual Recognition Arrangement
    \cite{std:bipm-cipm-mra}.
  \item The General Conference on Weights and Measures (CGPM),
    which adopts the SI's structural decisions. The CGPM is the
    political body whose 26th meeting in 2018 fixed the seven
    defining constants in their current form
    \cite{std:bipm-si2019}.
  \item National metrology institutes and designated institutes.
    NIST, NPL, PTB, NRC, NMIJ, NIM, VNIIM, CSIR-NPLI, and other
    bodies operate national or designated metrology functions; some
    countries use more than one designated institute for different
    fields. These institutes maintain realisations, participate in
    comparisons, and issue calibration certificates within published
    scopes \cite{std:bipm-cipm-mra}.
  \item Accredited calibration laboratories, assessed against
    ISO/IEC 17025 \cite{std:iso-iec-17025} by recognised accreditation
    bodies. These laboratories disseminate calibration services to
    industry and research within their accredited scopes.
  \item Working laboratories within manufacturers, research
    institutes, and service organisations, which calibrate their
    working instruments against secondary standards held by the
    accredited bodies or, where their volume justifies it, by the
    NMI directly.
\end{itemize}

\subsection{The CIPM Mutual Recognition Arrangement}

A manufacturer wants a calibration certificate issued under one
national metrology system to be intelligible and acceptable in
another. The CIPM Mutual Recognition Arrangement (CIPM MRA),
signed at S\`evres on 14 October 1999, supports that by recognising
NMI standards, calibration certificates, and published calibration
and measurement capabilities through comparisons, quality systems,
and peer review \cite{std:bipm-cipm-mra}.

Under the MRA, NMIs and designated institutes participate in periodic
key comparisons (international round-robin measurements of the same
quantity) whose results are published in the BIPM's KCDB (Key
Comparison Database). Legal or regulatory acceptance still depends
on the regulator, the accreditation scope, and the specific
requirement being applied; the MRA establishes the technical and
institutional basis for recognition without by itself overriding
domestic legal frameworks.

\subsection{ISO/IEC 17025 accreditation}

For laboratories that are not NMIs, the operational requirement is
accreditation under ISO/IEC 17025 \cite{std:iso-iec-17025}. The
standard specifies the management-system requirements (records,
calibration intervals, traceability documentation, personnel
competence, document control) and the technical requirements
(measurement uncertainty methodology, equipment, environmental
conditions, methods, sampling) that an accredited calibration
laboratory must meet.

ISO/IEC 17025 accreditation has direct and recurring cost:
application, assessment, corrective action, proficiency testing,
documentation, and surveillance. Assessment cycles are set by
accreditation bodies, so the cycle varies by jurisdiction. UKAS, for
example, uses annual surveillance with periodic full reassessment;
other accreditation bodies use different cycles. The accreditation
can be withdrawn when surveillance findings indicate the laboratory
has departed from the standard.

A reader who buys a calibration service from an accredited laboratory
acquires a certificate that names the accredited scope, the methods,
the references, and the uncertainty. A reader who buys a calibration
service outside an accredited scope must inspect the evidence
directly: method, reference standards, uncertainty budget,
competence, and traceability statement. Accreditation packages that
evidence in a recognised form; it does not replace the user's duty
to read the certificate.

\section{Calibration vs validation vs verification}\pathtag{standard}

Working engineering separates calibration, validation, and
verification because each answers a different audit question. The
vocabulary comes from VIM and quality-system standards
\cite{std:jcgm-vim}; sector-specific sources cover validation and
qualification at a level the general standards do not.

\engterm{Calibration} establishes a relationship between an
indication and a reference value. It is a measurement-system
operation, and we have already defined it above.

\engterm{Validation} confirms, through the provision of objective
evidence, that the requirements for a specific intended use or
application have been fulfilled. Validation is fitness-for-purpose:
it asks whether the measurement, the method, or the system is
suitable for its application. A method that is correctly calibrated
may still fail validation, because the calibrated quantity is not
the quantity the application actually requires. A pH probe
calibrated against pH-7 buffer is calibrated; whether it is
\emph{validated} for measuring the pH of a patient's blood depends
on whether the matrix interferences, the temperature compensation,
the response time, and the safety considerations have all been
documented to be acceptable for that use.

\engterm{Verification} provides objective evidence that specified
requirements have been fulfilled. Verification is conformance: it
asks whether a measurement, method, or system meets a specification.
Verification is mechanical relative to validation: a verification
procedure compares results against criteria, and produces a yes
or no.

The distinctions matter operationally. In a regulated environment
(medical devices, aerospace, nuclear, pharmaceuticals), an
instrument typically has a calibration certificate, a validation
record, and a verification log. Each plays a different role: the
calibration says how the instrument relates to the SI; the
validation says the calibrated instrument is suitable for the
application; the verification says the instrument continues to
meet specification under operational conditions.

A typical chain of evidence for a regulated measurement: the
manufacturer's calibration certificate provides traceability to the
SI; the user's validation report, drawing on the calibration
certificate plus application-specific tests, demonstrates fitness
for the intended use; the user's verification log, populated daily
or per-batch, demonstrates ongoing conformance.

\begin{mastery}
The fourth term, \engterm{qualification}, is used in
pharmaceutical and life-science contexts to mean the documented
demonstration that a system, equipment, or facility consistently
meets predetermined specifications. Qualification has substages
(installation qualification, operational qualification, performance
qualification) familiar from FDA's process-validation vocabulary.
The structure is parallel to the calibration-validation-verification
distinction; the vocabulary differs by industry. A reader working in
regulated industries learns the local terminology first.
\end{mastery}

\section{Drift, recalibration intervals, the cost of
calibration}\pathtag{core}

A calibration is a snapshot. Instruments drift after calibration:
electronic components age, mechanical bearings wear, optical
surfaces accumulate film, environmental exposure shifts component
parameters, packaging stresses re-distribute over time. Drift is the
mechanism that makes recalibration intervals necessary.

\subsection{Drift mechanisms}

Drift mechanisms vary by quantity and instrument type, but a few
patterns recur:

\begin{itemize}[leftmargin=*]
  \item \engterm{Component aging}: electronic components (resistors,
    capacitors, voltage references) drift on a logarithmic timescale
    after manufacturing. The first year typically sees the largest
    drift; subsequent years show progressively smaller annual
    drift, often parameterised as a fractional change per
    decade-of-age.
  \item \engterm{Mechanical wear}: bearings, contacts, springs, and
    sliding surfaces wear with use. The wear rate depends on duty
    cycle and operating environment; an instrument used twice a
    week wears at a different rate from one used continuously.
  \item \engterm{Surface contamination}: optical surfaces and
    high-precision mechanical surfaces accumulate films from
    handling, atmospheric deposition, and outgassing.
    Contamination usually shifts the calibration in a slowly
    monotonic direction until cleaning resets it.
  \item \engterm{Thermal cycling}: instruments stored or operated
    across a temperature range can experience permanent shifts as
    components transit phase-stress regimes. The effect is usually
    bounded but not always reversible.
  \item \engterm{Mechanical shock}: a dropped instrument or a
    shipping impact can produce a step shift in calibration that
    appears as a discrete event rather than gradual drift.
\end{itemize}

\subsection{Setting recalibration intervals}

The recalibration interval for an instrument is set by the rate at
which the instrument drifts beyond the acceptable tolerance for its
application. The procedure is, in principle, straightforward:

\begin{enumerate}[leftmargin=*]
  \item Calibrate the instrument repeatedly over an extended period
    without intermediate adjustment, recording each calibration's
    deviation from the previous reference.
  \item Estimate the rate of drift and its uncertainty from the
    history.
  \item Compute the interval at which the cumulative drift would
    reach the application's acceptable tolerance.
  \item Set the recalibration interval at a fraction of that
    computed interval, with the fraction chosen to give an
    acceptable risk of an instrument exceeding tolerance between
    calibrations.
\end{enumerate}

Many laboratories begin with a manufacturer recommendation, customer
requirement, regulatory requirement, or local default interval, then
adjust the interval using calibration history and risk. The general
rule is drift risk versus measurement stakes, not a universal
calendar interval; sourced examples should be cited by instrument
class and jurisdiction rather than asserted as universal practice.

\subsection{When statistical control suffices}

A laboratory that uses an instrument continuously has a different
option: statistical process control on the instrument's own output.
If the laboratory measures known control samples alongside its
working measurements and the control results stay within a defined
band, the instrument is operating within its calibration. If the
control results drift, the instrument is recalibrated regardless of
the calendar interval.

Statistical control reduces unnecessary recalibration on stable
instruments and triggers earlier-than-scheduled recalibration on
drifting ones. Control charts (Shewhart, EWMA, CUSUM) are the
operational tools; we take them up in chapter 8 of this volume on
statistics for engineers.

\subsection{The cost of calibration}

Calibration costs money. The actual price varies by accredited
status, scope, points measured, turnaround, on-site work, and
jurisdiction; a sourced 2026 price table from named providers is the
right reference rather than a generic range, and live prices are
better placed in a companion note than in the permanent text.

Annual recalibration of a working laboratory's instrument inventory
is a meaningful operating cost. Small laboratories may spend
modest amounts per year on calibration services; high-volume
laboratories spend orders of magnitude more; national-scale
metrology operations are larger still. The cost is one factor in
the recalibration-interval decision: too-frequent recalibration is
wasted money; too-infrequent recalibration risks unacceptable drift.
The interval that balances the two is, in practice, set by
regulatory requirement first and economic optimisation second.

\subsection{Computing the recalibration interval from drift data}

The textbook recalibration interval comes from three numbers: the
drift rate $d$ of the standard or instrument, the application's
acceptable out-of-tolerance probability $p$, and the original
calibration's standard uncertainty $u_{0}$. The international
calibration community has codified this calculation in two
reference documents: ILAC G24:2022
\cite{std:ch03t3-ilac-g24-2022}, which catalogues five interval
methods, and NCSLI RP-1 \cite{web:ch03t3-ncsli-rp-1-2010}, which is
the US calibration industry's standard practice. NASA's Reference
Publication 1342 \cite{std:ch03t3-nasa-rp-1342-1994} develops the
underlying statistical machinery: an interval-setting method based
on end-of-period reliability (EOPR), the fraction of instruments
that remain within tolerance at the end of the interval.

The simplest model: at calibration time, the standard's value is
known with uncertainty $u_{0}$. After elapsed time $t$, the
standard drifts at rate $d$ per unit time, with a small additional
random variation $\sigma_{r}$ from short-term noise. The total
deviation envelope at time $t$ is
\[
\delta(t) = d \cdot t + k \sqrt{u_{0}^{2} + \sigma_{r}^{2}\,},
\]
where $k$ is a coverage factor matched to the desired confidence
level ($k = 2$ for $95\%$, $k = 1.96$ for $95.0\%$ Gaussian, $k =
3$ for $99.7\%$). The recalibration interval is the value of $t$
at which the envelope reaches the application's tolerance $\Delta$:
\[
\tau = \frac{\Delta - k \sqrt{u_{0}^{2} + \sigma_{r}^{2}\,}}{d}.
\]
The interval is positive only when $k\sqrt{u_{0}^{2} +
\sigma_{r}^{2}} < \Delta$, that is, when the original calibration's
uncertainty already fits inside the tolerance with the required
margin. A user whose calibration uncertainty exceeds the tolerance
needs a better calibration, not a different interval.

A worked example, with named instruments. A Fluke 732C transfer
voltage standard has a manufacturer-specified drift rate $d =
\pm 2 \,\text{ppm}/\text{year}$ at $10\,\si{\volt}$ (current as of
the 732C's 2024 datasheet, descending from the original HP 732A
1976 specification). Suppose the application's tolerance is
$\Delta = 10 \,\text{ppm}$ and the original NMI calibration's
expanded uncertainty is $U_{0} = 1.5 \,\text{ppm}$ at $k = 2$
(typical for a 10\,V Josephson-traced NMI service). Then $u_{0} =
0.75 \,\text{ppm}$. Take $\sigma_{r} \approx 0.5 \,\text{ppm}$ for
short-term noise and operator variation. At $k = 2$ for $95\%$
confidence,
\[
\tau = \frac{10 - 2\sqrt{0.75^{2} + 0.5^{2}}}{2}
     = \frac{10 - 2 \cdot 0.90}{2}
     = \frac{8.2}{2}
     = 4.1 \,\text{yr}.
\]
A four-year interval is the upper bound. In practice, the laboratory
sets the interval at half to two-thirds of the bound (here $2$ to
$3 \,\text{yr}$) to allow for unexpected acceleration of drift, and
re-evaluates after each recalibration's as-found data: an instrument
that drifts faster than its spec reveals a shorter actual interval;
an instrument that drifts slower than its spec earns the right to a
longer one. NCSLI RP-1's adjustment rule
\cite{web:ch03t3-ncsli-rp-1-2010} formalises this feedback as a
geometric increase or decrease of the next interval by a fixed
ratio (typically $1.2\times$ for passes, $0.7\times$ for fails)
toward an EOPR target.

\subsection{Total cost of ownership for working instruments}

The recalibration interval is one variable in a larger ownership
decision. A working laboratory chooses, for each instrument,
between in-house calibration, an external accredited laboratory,
and the manufacturer's calibration service. Each option carries
distinct cost, turnaround, and operational-risk profiles. A
five-year total-cost-of-ownership (TCO) walkthrough for a single
instrument makes the trade-offs concrete.

Consider a Keysight 34470A 7.5-digit bench digital multimeter
\cite{web:ch03t3-keysight-34470a}, the workhorse DMM at the
\$4{,}500 list-price tier (current as of 2024). The instrument is
used continuously by an electrical-QC laboratory measuring DC
voltages, AC voltages, and resistances against industrial
production tolerances. Three calibration sourcing options:

\begin{itemize}[leftmargin=*]
  \item \engterm{In-house calibration} against a Fluke 5520A
    multifunction calibrator \cite{web:ch03t3-fluke-5520a} held by
    the laboratory. Capital cost: the 5520A list price is
    approximately \$22{,}000, plus \$3{,}000 for environmental
    enclosure and reference cabling. The 5520A itself must be sent
    out annually for accredited calibration at approximately
    \$1{,}800 per cycle. Per-DMM calibration cost: $\sim \$200$ of
    labour and amortised calibrator cost; turnaround under one day.
  \item \engterm{External accredited calibration} at a regional
    ISO/IEC 17025 \cite{std:iso-iec-17025} laboratory. Per-cycle
    cost: $\sim \$800$ for the 34470A's five-point DC voltage
    calibration with reported uncertainty budget. Turnaround:
    typically $10$--$15$ business days door-to-door for shipping
    and laboratory processing.
  \item \engterm{Vendor (manufacturer) calibration} returning the
    DMM to Keysight. Per-cycle cost: $\sim \$1{,}500$ for the
    one-year service. Turnaround: typically $4$--$6$ weeks
    door-to-door. Advantage: the manufacturer's report is signed
    against the manufacturer's specification.
\end{itemize}

The five-year TCO depends on the recalibration interval. Annual
recalibration is the cGMP-style default for medical and
pharmaceutical work; biennial intervals are typical for general
electronics QC. The two intervals lead to different TCOs by sourcing
option:

\begin{center}
\small
\begin{tabular}{lccc}
\toprule
Sourcing option & Annual recal. & Biennial recal. & Notes \\
\midrule
In-house (per DMM, 5-yr) & $\$1{,}000$ & $\$500$
  & Plus $\$25{,}000$ shared capital and $\$9{,}000$ shared
    5520A calibration \\
External accredited & $\$4{,}000$ & $\$2{,}000$
  & Shipping risk; lost production time \\
Vendor (Keysight) & $\$7{,}500$ & $\$3{,}750$
  & Longest turnaround; warranty alignment \\
\bottomrule
\end{tabular}
\end{center}

In-house is cheapest per DMM but only after the calibrator capital
is amortised. The break-even fleet size against an external
accredited laboratory is approximately $7$ DMMs at annual recal.
($25{,}000 / [(4{,}000 - 1{,}000) \cdot 5]$ rounded to whole DMMs);
below that fleet, external is cheaper, above it, in-house is. The
break-even shifts toward longer intervals because the capital
amortisation does not scale with interval.

The TCO table omits several real costs: shipping (often $\$50$ per
direction per DMM for accredited service); production downtime
during the calibration cycle (a 4-week vendor cycle implies a
shadow instrument or a delayed production schedule); the
opportunity cost of an in-house metrology technician's time (a
full-time technician at $\$120{,}000$ fully-loaded supports
approximately $400$ DMM calibrations per year at $\$300$ of
technician cost per calibration); and audit defendability (the
external and vendor reports carry the laboratory's accreditation
or the manufacturer's reputation as third-party evidence; the
in-house report depends on the laboratory's own accreditation
scope).

A working rule, current as of 2026: a small QC laboratory with
$< 5$ DMMs at biennial recal. uses an external accredited
laboratory; a medium laboratory with $5$--$20$ DMMs at annual recal.
benefits from in-house calibration if it already maintains broader
metrology infrastructure; a large fab with $> 100$ DMMs operates
in-house metrology as a profit centre and may sell calibration
services back to the local industry.

\subsection{Risk-scoring matrix for setting intervals across a fleet}

A laboratory with a heterogeneous instrument inventory chooses
intervals instrument-by-instrument by trading off three
dimensions, each scored on a $1$--$5$ scale. The product gives an
ordinal score that maps to an interval bucket:

\begin{center}
\small
\begin{tabular}{p{3.2cm}p{3.5cm}p{3.5cm}p{3.5cm}}
\toprule
Dimension & Low score (1) & Mid score (3) & High score (5) \\
\midrule
\emph{Criticality}
  & Internal-only QC, no regulatory exposure
  & Production QC affecting product release
  & Safety-critical, regulatory submission, accident-investigation
    defence \\
\emph{Drift rate}
  & $< 0.1$ of tolerance per year
  & $0.1$--$0.4$ of tolerance per year
  & $> 0.4$ of tolerance per year \\
\emph{Measurement frequency}
  & $< 100$ measurements per year
  & $100$--$1000$ per year
  & $> 1000$ per year, or continuous service \\
\bottomrule
\end{tabular}
\end{center}

The product of the three scores ranges from $1$ to $125$. A
working mapping to recalibration interval (the laboratory adjusts
these brackets to its own EOPR target and risk appetite):

\begin{itemize}[leftmargin=*]
  \item Product $\leq 8$: $36$-month interval. Low-stakes
    instruments used infrequently.
  \item Product $9$--$27$: $24$-month interval. The general working
    case.
  \item Product $28$--$60$: $12$-month interval. The cGMP-style
    default.
  \item Product $61$--$125$: $6$-month interval or shorter; consider
    in-service checking against a working standard between formal
    recalibrations.
\end{itemize}

A worked example. A pulse oximeter in a small clinic carries
criticality $5$ (clinical safety, FDA-regulated), drift rate $2$
(specified drift is $\sim 0.2\%$ SpO$_2$/year against a
$\pm 2\%$ tolerance), and measurement frequency $4$ (a few hundred
measurements per week). Product $= 40$; interval bucket $12$-month.
A general-purpose multimeter in an electrical-design bench
carries criticality $1$, drift rate $1$, frequency $3$;
product $= 3$; interval bucket $36$-month. A force gauge used for
acceptance testing of safety-critical fasteners in an aerospace
shop carries criticality $5$, drift rate $4$, frequency $5$;
product $= 100$; interval bucket $6$-month or in-service checking.
The matrix surfaces the rationale, prevents one-size-fits-all
calendar scheduling, and gives an audit trail explaining why a
specific interval was chosen.

\subsection{Worked fleet decision: 50-instrument contract
manufacturer}

A medium-sized contract manufacturer of automotive electronic
control units operates a metrology fleet of $50$ instruments
across electrical, mechanical, and thermal domains. The QC
manager applies the risk-scoring matrix to assign recalibration
intervals across the fleet and defends the resulting schedule in
an annual audit by a customer's Tier-1 supplier-quality team.

\textbf{Fleet inventory and scores.} The matrix is filled in for
each instrument; the table below shows the buckets and the
representative instruments in each bucket. Scores are
$C \times D \times F$ with $C$ criticality, $D$ drift, $F$
measurement frequency.

\begin{center}
\small
\begin{tabular}{p{2.5cm}p{1.5cm}p{1.0cm}p{7.5cm}}
\toprule
Bucket & Count & Interval & Representative instruments \\
\midrule
$\leq 8$ (low)
  & $14$
  & $36$\,mo
  & Lab DMMs on the prototyping bench (Keysight 34465A) used for
    rough engineering checks; reference voltmeters used only for
    occasional re-verification; spare oscilloscopes; demonstration
    instruments. \\
\addlinespace
$9$--$27$ (moderate)
  & $22$
  & $24$\,mo
  & Production DMMs on quality-inspection stations (Fluke 8845A);
    bench oscilloscopes on engineering benches; environmental
    chambers used outside their fastest-drift regions; torque
    drivers in subassembly stations; humidity loggers in long-term
    storage areas. \\
\addlinespace
$28$--$60$ (elevated)
  & $11$
  & $12$\,mo
  & Production-line scales (OIML class III at $\pm 0.5\,\si{\gram}$)
    used on every assembled board; in-circuit testers; HALT
    (highly-accelerated life test) chamber controllers; the
    plant's working Fluke 5520A multifunction calibrator; the
    Mettler XPR205 analytical balance \cite{web:ch03t3b-mettler-xpr}
    on the dimensional QC bench. \\
\addlinespace
$61$--$125$ (critical)
  & $3$
  & $6$\,mo or shorter
  & Safety-critical torque wrenches for airbag-detonator fastener
    installation (criticality $5$, drift $5$, frequency $5$,
    product $125$); the wirebond pull-tester gauge on the
    safety-critical microcontroller line ($5 \times 4 \times 5 =
    100$); the master comparator weight used to verify the
    production scales ($5 \times 3 \times 5 = 75$). All three
    are subject to in-service control-chart verification between
    formal recalibrations. \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Cost comparison.} The naive policy of recalibrating
every instrument annually would cost the fleet:
\[
50 \times \text{average } \$1{,}500/\text{instrument-year} +
\$50{,}000\,\text{shipping and handling} = \$125{,}000/\text{yr},
\]
or about $\$350{,}000/\text{yr}$ when the in-house technician
time, in-service control charts, audit preparation, and the
working-reference depreciation are added.

The differentiated policy from the risk matrix:
\begin{align*}
14 \text{ low-bucket} \times \$1{,}500 / 3\,\text{yr}
    &= \$7{,}000/\text{yr}, \\
22 \text{ moderate} \times \$1{,}500 / 2\,\text{yr}
    &= \$16{,}500/\text{yr}, \\
11 \text{ elevated} \times \$1{,}500 / 1\,\text{yr}
    &= \$16{,}500/\text{yr}, \\
3 \text{ critical} \times \$3{,}000 / 0.5\,\text{yr}\,
    \text{(includes shorter shipping cycle)}
    &= \$18{,}000/\text{yr}.
\end{align*}
Subtotal for outsourced calibration: $\$58{,}000/\text{yr}$. Add
$\$10{,}000/\text{yr}$ in shipping and handling (the longer
intervals mean fewer shipments), $\$80{,}000/\text{yr}$ in
in-house technician time (two technicians at half-time loaded
$\$80{,}000/\text{technician-year}$), $\$30{,}000/\text{yr}$ in
working-reference depreciation, $\$15{,}000/\text{yr}$ in
in-service control-chart and audit-preparation effort, and
$\$10{,}000/\text{yr}$ in software (calibration-management
system, control-chart software, document-control system). Total:
$\$203{,}000/\text{yr}$, or about $40\%$ below the naive
$\$350{,}000/\text{yr}$ baseline.

\textbf{Audit defence.} The QC manager defends the schedule by
producing, for each instrument:
\begin{enumerate}[leftmargin=*]
  \item The filled-in score sheet showing $C$, $D$, $F$ and the
    resulting product.
  \item The drift evidence: either the manufacturer's stated
    drift specification, the laboratory's own historical as-found
    data showing actual drift below the assumed value, or both.
  \item The criticality justification, tied to the instrument's
    role in the customer's PPAP (production-part approval
    process) or in safety-relevant fastener torque, where
    applicable.
  \item The in-service verification plan for any instrument
    whose interval exceeds one year (a $24$- or $36$-month
    interval requires intermediate checks against a transfer
    standard or against an internal stability-check artefact at
    least quarterly).
  \item The escalation rule: any as-found result outside $80\%$
    of the manufacturer specification triggers an immediate
    interval halving for the next cycle.
\end{enumerate}
A customer's Tier-1 supplier-quality auditor accepts the
schedule when the defence is documented, the as-found history
supports the assumed drift rates, and the escalation rule has
been exercised at least once (showing the system actually
detects and responds to drift). A schedule defended this way
also survives a third-party ISO/IEC 17025 surveillance audit and
an IATF 16949 process audit on the customer's behalf.

\textbf{The 40\% saving is conditional.} The differentiated policy
returns $\$147{,}000/\text{yr}$ only if the as-found data
continues to support the interval extensions. A regression of
the policy occurs when (a) a low-bucket instrument enters
production-QC service without rescoring (the most common audit
finding), (b) a drift rate quietly accelerates (battery-backed
real-time clocks past $10$ years; capacitive humidity sensors
past $5$ years are typical), or (c) the customer's requirements
tighten and what was elevated becomes critical. The QC manager
rescore the fleet annually as a discipline; without the
rescoring, the saving is borrowed against future
out-of-tolerance findings.

\subsection{Type A versus Type B uncertainty: when each applies}

The GUM \cite{std:ch03t3-jcgm-100-2008} divides uncertainty
contributions by how they are estimated, not by their physical
nature. Type A contributions are estimated by statistical analysis
of a series of repeated observations; Type B contributions are
estimated by any other means: manufacturer specifications,
calibration certificates, handbook data, judgement. The same
physical effect can be a Type A contribution in one budget and a
Type B contribution in another, depending on whether the analyst
has repeated measurements that probe it.

The decision tree for a working budget:

\begin{enumerate}[leftmargin=*]
  \item Identify the contribution. (A thermal-EMF in a cable, a
    temperature variation, a resolution limit, an operator bias.)
  \item Ask: does the laboratory have $n \geq 5$ repeated
    measurements of this contribution under conditions matching
    the working use?
  \item If yes: estimate Type A by sample standard deviation
    $s$; the standard uncertainty of the mean of $n$ measurements
    is $u = s/\sqrt{n}$ (GUM Section 4.2).
  \item If no: estimate Type B from the manufacturer specification,
    calibration certificate, or judgement. Choose a distribution
    consistent with what is known (GUM Section 4.3).
\end{enumerate}

The distribution choices for Type B follow a small vocabulary:

\begin{itemize}[leftmargin=*]
  \item Manufacturer accuracy spec stated as $\pm a$ (no
    distribution claimed): treat as a uniform distribution on
    $[-a, +a]$. The standard uncertainty is $u = a/\sqrt{3}$.
    This is the maximum-entropy choice consistent with knowing
    only the worst-case bound.
  \item Manufacturer accuracy spec stated as $\pm a$ at $k = 2$
    (Gaussian implied): the standard uncertainty is $u = a/2$.
    Read the datasheet's footnotes to determine which.
  \item Calibration certificate reporting expanded uncertainty
    $U$ at coverage factor $k$: the standard uncertainty is
    $u = U/k$.
  \item Resolution limit (a digital display reading to the nearest
    unit $R$): treat as a uniform distribution on $[-R/2, +R/2]$;
    standard uncertainty $u = R/(2\sqrt{3})$.
  \item Triangular distribution on $[-a, +a]$ (for an effect
    expected to peak at zero): $u = a/\sqrt{6}$.
\end{itemize}

The sample-size calculation for Type A determines the minimum
number of repeated measurements that brings the random
contribution below a target. To halve the Type A contribution at
fixed sample standard deviation, the analyst quadruples $n$. The
practical minimum is typically $n = 5$ readings (giving Student's
$t$-distribution coverage at usable confidence); the marginal value
of additional readings declines after $n = 20$. For instruments
operating near a regulatory tolerance, $n = 10$ to $n = 30$
repeated readings under matched conditions is the working
guideline. Below $n = 5$, the analyst falls back to a Type B
estimate from manufacturer specification.

A worked vignette. A laboratory is budgeting the uncertainty of a
$10\,\si{\volt}$ DC measurement. The DMM's resolution is
$1\,\si{\micro\volt}$ (Type B: $u_{\text{res}} = 0.5/\sqrt{3} =
0.29 \,\si{\micro\volt}$). The DMM's manufacturer spec at one year
is $\pm 16\,\si{ppm}$ of reading $\pm 2\,\si{ppm}$ of range
\cite{web:ch03t3-keysight-34470a}, which at $10\,\si{\volt}$ on
the $10\,\si{\volt}$ range works to $\pm (160 + 20) =
\pm 180\,\si{\micro\volt}$. Treated as a uniform distribution
without further information, $u_{\text{spec}} = 180/\sqrt{3} =
104 \,\si{\micro\volt}$. The laboratory then measures the same
DMM against its working reference five times, computing a sample
standard deviation $s = 8\,\si{\micro\volt}$ of the readings, so
$u_{\text{rep}} = 8/\sqrt{5} = 3.6 \,\si{\micro\volt}$ (Type A).
The combined uncertainty is dominated by the Type B manufacturer
spec; bringing the manufacturer spec into the budget at all
acknowledges that the laboratory has not characterised the DMM
beyond the manufacturer's claim. A laboratory that performs its
own annual calibration replaces the manufacturer Type B with the
calibration certificate's smaller uncertainty, possibly an order
of magnitude smaller, and the random Type A contribution becomes
the dominant term.

\subsection{An end-to-end QC lab worked example}

A small electrical-QC laboratory in a regional electronics
manufacturer operates four reference instruments, each with its
own calibration cadence and sourcing decision. The inventory and
the year-one cost summary illustrate the integrated decision:

\begin{itemize}[leftmargin=*]
  \item Keysight 34470A 7.5-digit DMM
    \cite{web:ch03t3-keysight-34470a}. Used for incoming inspection
    of voltage references and resistance standards.
    Criticality $3$ (production QC), drift $2$, frequency $5$;
    product $30$, $12$-month interval. Sourced externally; cost
    $\$800$/yr.
  \item Fluke 5520A multifunction calibrator
    \cite{web:ch03t3-fluke-5520a}. Calibrates customer DMMs and
    internal DMMs. Criticality $5$ (the laboratory's metrological
    anchor), drift $1$ (5520A drift is $\leq 10\,\si{ppm}$/yr at
    $10\,\si{\volt}$), frequency $4$;
    product $20$, $12$-month interval (per accreditation
    requirements). Sourced from Fluke's accredited cal lab; cost
    $\$1{,}800$/yr including shipping.
  \item Fluke 1551A ``Stik'' thermometer
    \cite{web:ch03t3-fluke-1551a-stik}. Reference for the
    laboratory's temperature-controlled calibration enclosure.
    Criticality $3$, drift $2$, frequency $3$; product $18$,
    $24$-month interval. Sourced externally; cost $\$400$ amortised
    over two years $= \$200$/yr.
  \item B\&K Precision 4054C function generator
    \cite{web:ch03t3-bk-4054c}. Used to verify customer
    oscilloscopes and frequency counters. Criticality $2$,
    drift $2$, frequency $3$; product $12$, $24$-month interval.
    Sourced externally; cost $\$300$ amortised $= \$150$/yr.
\end{itemize}

Annual calibration cost summary, year one: $\$800 + \$1{,}800 +
\$200 + \$150 = \$2{,}950$. To this add shipping and packaging
($\sim \$250$/yr), the laboratory's own in-service control charts
on the 5520A (one technician-day per month $\sim \$3{,}600$/yr),
and the cost of the formal annual review of the calibration
schedule against the prior year's as-found data ($\sim \$1{,}500$
in technician and supervisor time). Total year-one cost of
running the calibration discipline: $\sim \$8{,}300$. The
laboratory's customer-facing revenue from calibration services
(if any) and the avoided cost of a single out-of-tolerance event
on a customer-released batch (typically $\$50{,}000$ to
$\$500{,}000$ depending on the customer's downstream exposure)
are the framing numbers that justify the expenditure.

The acceptable risk of an out-of-tolerance discovery, in the
laboratory's quality manual, is typically $\leq 5\%$ EOPR for
production-QC instruments and $\leq 1\%$ for safety-critical
work. The interval-setting calculation works backward from those
targets through the drift rate and the calibration uncertainty
to determine the interval; a year of as-found data either
confirms the choice or triggers an adjustment per NCSLI RP-1
\cite{web:ch03t3-ncsli-rp-1-2010}. The end-of-year audit
trail consists of the calibration certificates, the as-found
data, the in-service control charts, the recalibration adjustments
made under the geometric-adjustment rule, and the documented
disposition of any out-of-tolerance findings. A laboratory whose
auditor cannot reconstruct the discipline from the audit trail
has not done the discipline.

\subsubsection*{Working standards purchased to support the
schedule}

The four customer-facing instruments depend on a small set of
in-house working standards. The full standards inventory and
their purchase cost is:
\begin{itemize}[leftmargin=*]
  \item \textbf{Fluke 7526A precision process calibrator}
    \cite{web:ch03t3b-fluke-7526a}, purchased for $\$5{,}800$.
    Used as the in-house check standard against the 5520A in the
    six months between accredited recalibrations. The 7526A's
    DC voltage accuracy of $\pm 0.005\%$ at $10\,\si{\volt}$
    gives a $\pm 500\,\si{\micro\volt}$ check uncertainty against
    the 5520A's $\pm 18\,\si{\micro\volt}$ specification, a
    $28{:}1$ ratio that comfortably detects any drift in the
    5520A's output.
  \item \textbf{Hart Scientific 5618 secondary-standard PRT}
    \cite{web:ch03t3b-hart-5618}, purchased for $\$3{,}200$, with
    triple-point-of-water cell ($\$2{,}500$) and resistance bridge
    ($\$8{,}500$). The PRT's calibration uncertainty of
    $\pm 0.006\,\si{\kelvin}$ at the triple-point of water is the
    laboratory's temperature anchor; the in-house Fluke 1551A
    Stik thermometers are checked against this PRT between their
    external annual recalibrations.
  \item \textbf{Class E2 mass set, $1\,\si{\milli\gram}$ to
    $1\,\si{\kilo\gram}$, OIML R 111 compliant}, purchased for
    $\$4{,}200$. Used to verify the Mettler XPR205
    \cite{web:ch03t3b-mettler-xpr} on the dimensional QC bench
    quarterly and to verify a Keysight 3458A
    \cite{web:ch03t3b-keysight-3458a} on the production-QC bench
    against the 5520A monthly.
  \item \textbf{Working-reference Zener voltage standard} (Fluke
    732B equivalent), purchased used for $\$3{,}500$. Provides a
    portable $10\,\si{\volt}$ reference that is shipped to the
    accredited calibration laboratory annually as the
    laboratory's transfer artefact. The 732B's documented drift of
    $\pm 2\,\si{ppm}/$year provides the laboratory's stability
    history independent of the 5520A.
  \item \textbf{Working-reference resistance standards}
    ($1\,\si{\ohm}$, $10\,\si{\ohm}$, $100\,\si{\ohm}$,
    $1\,\si{\kilo\ohm}$, $10\,\si{\kilo\ohm}$), purchased as a
    matched set for $\$4{,}500$. Used to verify DMM resistance
    accuracy across decades.
\end{itemize}
Total in-house standards investment: $\$32{,}200$. Depreciated
over a $10$-year working life: $\$3{,}220/\text{yr}$.

\subsubsection*{Recalibration schedule, year by year}

The schedule has external (accredited) and internal (in-house)
events:
\begin{enumerate}[leftmargin=*]
  \item \textbf{January.} The Fluke 5520A
    \cite{web:ch03t3-fluke-5520a} is shipped to Fluke's
    A2LA-accredited calibration laboratory in Everett, WA, with
    a stated $10$-business-day turnaround. The laboratory loans
    a spare 5520A from its rental fleet to cover the gap
    (rental: $\$2{,}500$/month including shipping).
  \item \textbf{January-February.} The Hart Scientific 5618 PRT
    \cite{web:ch03t3b-hart-5618} is shipped to NIST's calibration
    services lab through the NIST CSP (Calibration Services
    Program) for ITS-90 fixed-point calibration; turnaround
    typically $6$-$8$ weeks. The laboratory's Fluke 1551A units
    are checked against the in-house PRT on its return.
  \item \textbf{April.} Mid-year in-house verification of the
    34470A \cite{web:ch03t3-keysight-34470a} against the 5520A,
    documented on a standard verification form. Any drift greater
    than $25\%$ of specification triggers an out-of-cycle
    accredited recalibration.
  \item \textbf{July.} Working-reference Zener (Fluke 732B
    equivalent) shipped to A2LA-accredited laboratory (Transcat,
    Inc.\ in this case \cite{web:ch03t3b-transcat}) for annual
    recalibration. Turnaround typically $5$-$10$ business days.
  \item \textbf{October.} Mid-year in-house verification of the
    34470A repeated. Mass set verified against the Class E1
    weights at the accredited mass laboratory annually (the
    laboratory chose a smaller A2LA-only lab for this:
    $\$600$/yr).
  \item \textbf{December.} End-of-year review. The QC manager
    consolidates the year's as-found data into a single
    spreadsheet, computes the laboratory's overall
    end-of-period reliability (EOPR), and triggers interval
    adjustments per NCSLI RP-1 \cite{web:ch03t3-ncsli-rp-1-2010}
    for the next year.
\end{enumerate}

\subsubsection*{Sourcing comparison: send-out options}

The 5520A and the 5618 are the two highest-stakes send-outs.
The laboratory ran a sourcing comparison in 2024:
\begin{center}
\small
\begin{tabular}{p{3.0cm}p{3.2cm}p{3.0cm}p{4.0cm}}
\toprule
Option & Cost (5520A) & Turnaround & Note \\
\midrule
NIST CSP direct
  & $\$3{,}500$
  & $8$-$12$\,wk
  & Highest pedigree; longest queue; impractical for the
    laboratory's $10$-business-day target. \\
Fluke A2LA cal lab
  & $\$1{,}800$
  & $10$\,bd
  & Manufacturer's own A2LA-accredited service; chosen for the
    5520A. \\
Transcat (commercial)
  & $\$1{,}500$
  & $5$-$10$\,bd
  & A2LA-accredited; chosen for the Zener reference. \\
Independent A2LA-only lab
  & $\$1{,}200$
  & $7$-$14$\,bd
  & Lowest cost; chosen for the mass set; reviewed annually for
    continued accreditation scope. \\
\bottomrule
\end{tabular}
\end{center}
The laboratory's policy is to use the manufacturer's own
accredited service for the metrological anchor (the 5520A) and a
commercial A2LA-accredited lab for the working transfers. A
direct-NIST send-out is reserved for the periodic re-anchoring
of the in-house Hart 5618 PRT, where NIST's traceability
pedigree is the working reason for the choice.

\subsubsection*{Year-one operating cost, fully loaded}

The full year-one cost of the calibration program, including all
the working standards depreciation, in-house technician effort,
software, and consumables:
\begin{align*}
\text{External calibrations (customer-facing inventory)} &= \$2{,}950, \\
\text{External calibrations (working standards)} &= \$3{,}100, \\
\text{Hart 5618 PRT to NIST CSP, biennial allocation} &= \$1{,}500, \\
\text{Shipping and handling, all send-outs} &= \$1{,}200, \\
\text{Loaner-5520A rental} &= \$2{,}500, \\
\text{In-house technician time (0.4 FTE)} &= \$36{,}000, \\
\text{Quality-system maintenance (procedures, audits)} &= \$8{,}000, \\
\text{In-service control charts and review} &= \$5{,}000, \\
\text{Software (calibration management, control-chart)} &= \$3{,}500, \\
\text{In-house standards depreciation} &= \$3{,}220, \\
\text{Environmental control (chamber, humidity)} &= \$2{,}500.
\end{align*}
Total year-one operating cost: $\$69{,}470 \approx \$70{,}000/
\text{yr}$ for a four-customer-instrument lab supported by five
in-house working standards. A typical small accredited
calibration laboratory operating at this scale charges its
customers $\$120{,}000$ to $\$180{,}000/\text{yr}$ in calibration
services to cover the program, generating a margin sufficient to
fund the next investment cycle (replacement of the 5520A at
$\sim$year~$10$, replacement of the Hart 5618 at $\sim$year~$15$).
A common back-of-envelope figure of $\sim\$48{,}000/\text{yr}$
for the same lab captures the external services alone ($\$8{,}550
/\text{yr}$) plus a half-loaded technician and software, and
understates the full program by the depreciation, the in-house
standards investment, and the quality-system overhead, all of
which the auditor expects to see costed.

\subsubsection*{Acceptable risk and the EOPR target}

The QC laboratory's quality manual fixes a target end-of-period
reliability (EOPR) for each instrument tier. A representative
target structure:
\begin{itemize}[leftmargin=*]
  \item Safety-critical instruments (the 5520A as metrological
    anchor; the master comparator weight for production scales):
    EOPR $\geq 99\%$, i.e.\ the probability of finding the
    instrument out of tolerance at its next calibration is
    $\leq 1\%/\text{interval}$.
  \item Production-QC instruments (the 34470A, the 1551A): EOPR
    $\geq 95\%$.
  \item Non-critical instruments (the 4054C function generator):
    EOPR $\geq 90\%$.
\end{itemize}
For the 5520A with a documented drift rate
$\sigma_{d} = 5\,\si{ppm}/\sqrt{\text{year}}$ in the $10\,\si{
\volt}$ DC output and a manufacturer specification tolerance
$T_{s} = 18\,\si{ppm}$ at one year, the laboratory computes the
interval $\tau$ from the Gaussian tail:
\[
\Pr[|d(\tau)| > T_{s}] = 2 \Phi(-T_{s}/(\sigma_{d}\sqrt{\tau}))
\leq 0.01,
\]
where $\Phi$ is the standard normal CDF. Solving for $\tau$:
\[
T_{s}/(\sigma_{d}\sqrt{\tau}) \geq z_{0.005} = 2.576,
\quad \sqrt{\tau} \leq T_{s}/(2.576\,\sigma_{d}),
\quad \tau \leq (18/(2.576 \times 5))^{2} = 1.95\,\text{yr}.
\]
A $12$-month interval easily meets the $1\%$ target. A
$24$-month interval (often proposed to save shipping cost) gives
$\sqrt{2} \times 5 = 7.07\,\si{ppm}/\text{interval}$ standard
deviation and a tail probability of $\Pr[|d| > 18] = 2\Phi(-2.55)
= 1.1\%$, just outside the safety-critical target. The laboratory
keeps the 12-month interval and uses the in-service control
charts to confirm that the actual as-found drift remains within
the assumed $5\,\si{ppm}/\sqrt{\text{year}}$ envelope.

For the 4054C with $\sigma_{d} = 50\,\si{ppm}/\sqrt{\text{year}}$
in the $1\,\si{\kilo\hertz}$ output and a tolerance $T_{s} =
500\,\si{ppm}$, the $10\%$ EOPR target allows
\[
\tau \leq (500/(z_{0.05} \cdot 50))^{2} = (500/(1.645 \times 50))^{2}
= 36.9\,\text{yr},
\]
far beyond any practical recalibration interval. The drift is so
small relative to the tolerance that the $24$-month interval is
chosen on operational grounds (consistency with other
non-critical instruments, simplicity of scheduling) rather than
on a strict EOPR calculation. The asymmetry between the 5520A's
tight EOPR-driven interval and the 4054C's slack
operationally-driven interval is the differentiated discipline:
each instrument's interval comes from the trade-off appropriate
to its tier.

The EOPR computation hides three traps. First, the assumed
$\sigma_{d}$ is the standard deviation of drift, which the
laboratory measures only by accumulating multiple cycles of
as-found data; year-one calculations use the manufacturer's
specification of drift as a Type B estimate, and refinement
happens only after the laboratory has its own history. Second,
the Gaussian assumption breaks for instruments whose drift is
monotonic rather than random-walk (a battery-backed Zener
reference drifts monotonically with internal-element ageing for
the first $1{,}000$ hours, then more nearly random-walk); a
monotonic drift requires a different model (often a linear
trend with the random-walk superimposed). Third, the tolerance
$T_{s}$ is the manufacturer specification, but the laboratory's
working tolerance for in-house verification may be tighter; the
EOPR calculation must use the tolerance the laboratory actually
holds against.

\subsubsection*{Reading the calibration certificate}

A reader presented with an unfamiliar accredited calibration
certificate looks for, in the order they appear under
ISO/IEC 17025 Clause 7.8.4
\cite{std:ch03t3b-iso17025-clause7-8}:
\begin{enumerate}[leftmargin=*]
  \item \textbf{Title and identification.} ``Calibration
    Certificate'' (the phrase is mandated by the standard), the
    certificate number, the issuing laboratory's name and
    address, the laboratory's accreditation body and certificate
    number (e.g., ``A2LA Accredited Certificate No.\ 1234.01'').
    The presence of the accreditation logo without the
    accompanying certificate number is a flag.
  \item \textbf{Customer and instrument.} Customer name and
    address, the instrument's manufacturer, model number, serial
    number, and customer asset number. Mismatch between the
    asset number on the certificate and the asset number on the
    instrument's calibration sticker is the most common audit
    finding in non-accredited shops.
  \item \textbf{Date received, date calibrated, date issued.} The
    received-and-calibrated dates determine the as-found data's
    relevance to the prior interval's drift estimate. A
    certificate dated three weeks after calibration is a flag for
    delayed reporting.
  \item \textbf{Method and reference standards.} The procedure
    name and revision (e.g., ``per Fluke 5520A Operator's Manual
    Section 6, Rev D, 2020''), and the laboratory's reference
    standards used, identified by manufacturer, model, serial
    number, the higher-level certificate number, and the
    higher-level laboratory.
  \item \textbf{Environmental conditions.} Temperature ($23 \pm
    1\,\si{\degreeCelsius}$ is the metrological default),
    relative humidity ($40$-$60\%$), barometric pressure if
    relevant, AC mains stability. Missing environmental data
    means the calibration is not reproducible.
  \item \textbf{Results table.} For each test point: the
    reference value, the unit-under-test indication, the
    correction or error, the expanded uncertainty $U$ at the
    stated coverage factor $k$ (almost always $k = 2$ for
    accredited certificates per ILAC P14
    \cite{std:ch03t3b-ilac-p14}), and, optionally, the
    as-found and as-left values if an adjustment occurred during
    the calibration.
  \item \textbf{Statement of compliance and decision rule.} A
    sentence indicating whether the instrument complies with a
    stated specification, with the decision rule explicitly named
    (e.g., ``Test Uncertainty Ratio (TUR) $\geq 4{:}1$,
    conformity decision per ILAC G8 simple acceptance''). A
    statement of compliance without a decision rule is the third
    most common audit finding.
  \item \textbf{Signatures and traceability statement.} The
    technician and approver names and signatures, and a statement
    that all measurements are traceable to the SI through a named
    NMI (NIST, NPL, PTB, etc.). The traceability statement is the
    legal hook for the chain.
\end{enumerate}
Missing fields signal where the certificate's traceability
breaks. A certificate without a coverage factor, without a
reference-standard identification, or without a temperature is
not an accredited calibration certificate, regardless of what
its title page says. A reader who cannot answer ``to what NMI,
through what chain, with what uncertainty?'' from the
certificate's face has been handed a piece of paper, not a
calibration.

\begin{estimation}
\textbf{What does it cost to operate a small pressure-calibration
laboratory at A2LA-accredited quality?}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A small pressure laboratory mirrors the electrical-QC laboratory's
structure with pressure-specific working standards. The capital
investment: a Fluke 2700G portable reference at
$\sim\$15{,}000$ \cite{web:ch03t3b-fluke-2700g}, a WIKA CPB5000
hydraulic pressure balance at $\sim\$45{,}000$
\cite{web:ch03t3b-wika-cpb5000}, a Mensor CPC6050 digital
reference manometer at $\sim\$25{,}000$
\cite{web:ch03t3b-mensor-cpc6050}, a temperature-controlled
calibration room at $\sim\$30{,}000$ in build-out, and the
$\sim\$50{,}000$ of ancillary equipment (mass sets, calibration
tubing, gauge fittings, pressure media): total capital
investment $\sim\$165{,}000$. Annual external calibrations: the
CPB5000 to NIST CSP or a top-tier accredited lab at
$\sim\$4{,}000$/yr (the WIKA factory does not provide A2LA-scope
service for this model), the 2700G at $\sim\$1{,}500$/yr to
Fluke's A2LA-accredited service, the CPC6050 at $\sim\$1{,}200$/yr
to a commercial A2LA lab, mass set at $\sim\$800$/yr, ancillary
gauges at $\sim\$2{,}000$/yr: $\sim\$9{,}500$/yr in external
services. Add one full-time pressure technician at
$\sim\$110{,}000$/yr fully loaded, $\sim\$15{,}000$/yr in
quality-system effort and software, $\sim\$8{,}000$/yr in
consumables (calibration oil, leak fittings, replacement
o-rings, helium for the gas-bearing piston), $\sim\$17{,}000$/yr
in capital depreciation (over $10$-year working life), and
$\sim\$5{,}000$/yr in environmental control: total annual
operating cost $\sim\$165{,}000$/yr. At a typical billing rate of
$\sim\$250$-$\$400$/hour for accredited pressure calibration, the
laboratory needs $\sim\$220{,}000$-$\$330{,}000$ in revenue to
operate profitably, implying $\sim 700$-$1000$ billable hours per
year. A small pressure laboratory operating below this threshold
either subsidises its calibrations from an internal cost centre
(a refinery, a power plant, an aerospace shop with its own
metrology budget) or operates non-accredited. The accreditation
overhead is a fixed cost; the working pressure laboratory either
runs at scale or runs as an internal-only function.
\end{estimation}

\begin{estimation}
\textbf{What is the recalibration interval for a kitchen scale
used to weigh meal portions to within $\pm 5 \,\si{\gram}$?}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A typical household kitchen scale has a manufacturer-claimed
specification around $\pm 1 \,\si{\gram}$ at $100 \,\si{\gram}$ and
$\pm 5 \,\si{\gram}$ at $1 \,\si{\kilogram}$. Drift mechanisms for a
strain-gauge kitchen scale include load-cell creep, electronics
aging, and zero-point drift from temperature and humidity exposure.
Manufacturer specifications typically state initial accuracy without
naming a drift rate.

A working estimate starts from the user's tolerance, not from the
scale's display resolution. If the application allows
$\pm 5 \,\si{\gram}$ and the scale repeats within about
$\pm 1 \,\si{\gram}$ under a local check, the immediate risk is
small. A yearly check with two or three known masses is usually
more defensible than asserting that a household scale follows a
formal recalibration interval. If the check shows a bias near the
tolerance, the scale is no longer fit for that use.

If the application requires $\pm 1 \,\si{\gram}$, the tolerance is
the same size as the scale's likely repeatability, and the correct
engineering answer is a better instrument or a lower-stakes
measurement. The headroom-to-tolerance ratio is the lever;
resolution requirements above the instrument's specification cannot
be reached by recalibration alone.
\end{estimation}

\input{volumes/01-quantity/ch03-calibration-and-traceability/figures/fig-drift-recalibration.tex}

\begin{estimation}
\textbf{What is the calibration-and-recalibration cost for a
mid-sized US semiconductor fab's metrology fleet?}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A mid-sized 300\,mm semiconductor fab operates approximately
$10{,}000$ instruments (mass-flow controllers, pressure
transducers, temperature probes, voltage references, leak
detectors, particle counters, ellipsometers, profilometers,
thickness gauges, alignment systems). Annual calibration cost per
instrument averages $\$200$-$\$2000$ depending on tier; assume a
weighted average of $\$500$. Total annual calibration cost: $\$5
\,\text{million}/\text{yr}$. Add the in-house metrology workforce
($\sim 50$ engineers at $\$150{,}000$ each fully loaded = $\$7.5
\,\text{million}/\text{yr}$) and the reference-equipment depreciation
($\sim \$2 \,\text{million}/\text{yr}$ amortised over equipment
life): total metrology operating cost $\approx \$15 \,\text{million}/
\text{yr}$.

The fab's annual revenue is typically $\$5$-$\$15 \,\text{billion}$
for a single 300\,mm site; the metrology cost is $\sim 0.1$-$0.3\,\%$
of revenue. This appears small relative to revenue but is the
denominator of the yield calculation: a $1\,\%$ improvement in
yield from better measurement traceability returns $\$50$-$\$150
\,\text{million}/\text{yr}$. The cost-benefit overwhelmingly favours
the metrology investment, and is one reason the semiconductor
industry has been a leading consumer of metrology services.
\end{estimation}

\section{Cross-domain examples: voltage, mass, length, frequency,
temperature, pressure}\pathtag{standard}

At the primary-standard end, each quantity has its own apparatus,
comparison programme, and uncertainty budget. Six examples show how
different the realisations are.

\subsection{Voltage}

The volt is realised through the AC Josephson effect: a
superconductor-insulator-superconductor (SIS) tunnel junction
biased with a stable microwave field develops a DC voltage
$V_{n} = n f/K_{J}$, where $f$ is the microwave frequency (typically
$70$ to $96\,\si{\giga\hertz}$ in working arrays), $n$ is an
integer counted from the Shapiro-step structure of the
current-voltage characteristic, and $K_{J} = 2e/h \approx
483\,597.848\,4 \,\si{\giga\hertz\per\volt}$ is the Josephson
constant. The post-2019 SI fixes $e$ and $h$ by definition, so
$K_{J}$ is exact and the realised voltage uncertainty inherits
the microwave-frequency uncertainty alone (typically $10^{-12}$
fractional from a hydrogen-maser-locked source).

A working Josephson voltage standard array contains $10^{4}$ to
$3 \times 10^{5}$ junctions stacked in series on a silicon
substrate cooled to $4.2\,\si{\kelvin}$ in liquid helium. The
junctions are fabricated as thin-film Nb-AlO$_{x}$-Nb structures
in microstrip transmission lines so the microwave field reaches
each junction at the same phase. At $f = 75\,\si{\giga\hertz}$
and $n = 1$ per junction, $10^{4}$ junctions deliver
$V_{\text{array}} = 10\,\si{\volt}$ to within parts in $10^{10}$
of the calculated value, set by the system uncertainty (microwave-
frequency error, parasitic-DC leakage, thermal-EMF in the
twisted-pair output lines). NIST's Programmable Josephson
Voltage Standard (PJVS) at Gaithersburg outputs $-10\,\si{\volt}$
to $+10\,\si{\volt}$ programmable in $1\,\si{\micro\volt}$ steps
with a per-step uncertainty below $1\,\si{\nano\volt}$ ($\sim
10^{-10}$ fractional), current as of 2026 \cite{web:nist-pjvs}.

The dissemination chain. A secondary voltage standard, typically
a Zener-diode reference such as the Fluke 732C (a thermally-
controlled buried-Zener architecture descended from the original
Hewlett-Packard 732A of 1976), holds the volt at $\pm 2\,\si{ppm}/
\text{year}$ stability when transported. The accredited laboratory's
Zener references are compared against the laboratory's Josephson
array twice a year and against the customer's reference at the
laboratory's accredited measurement capability (CMC). Typical CMC
values from accredited US laboratories at $10\,\si{\volt}$ DC are
$\pm 0.6\,\si{ppm}$ at $k = 2$, derived from a budget that allocates
$\pm 0.3\,\si{ppm}$ to the Zener short-term stability, $\pm 0.3\,
\si{ppm}$ to the Josephson-Zener comparison, $\pm 0.2\,\si{ppm}$
to environmental and operator effects, and $\pm 0.1\,\si{ppm}$ to
the customer-laboratory transfer.

The working laboratory's digital multimeter (DMM) is the final
link. A 7.5-digit bench DMM such as the Keysight 3458A (introduced
1989, still the working metrology DMM) reports a
$24$-hour accuracy specification at $10\,\si{\volt}$ DC of
$\pm 8\,\si{ppm}$ of reading $\pm 0.5\,\si{ppm}$ of range at
$\pm 1\,\si{\celsius}$ from calibration temperature; the same
instrument's $90$-day specification is $\pm 14\,\si{ppm}$ and
$1$-year is $\pm 18\,\si{ppm}$ at the same conditions. The
8.5-digit Keysight 3458A or the equivalent Fluke 8588A reaches
parts-per-billion accuracy when properly conditioned and
calibrated against the laboratory's reference Zener.

The voltage traceability chain has the shortest dissemination
chain of the working SI quantities because it rests on two
exact-by-definition constants and a single quantum-mechanical
resonance condition; the chain for resistance (through the quantum
Hall effect, $R_{K} = h/e^{2} \approx 25\,812.807\,4 \,\si{\ohm}$)
and for current (through the voltage standard and the resistance
standard via Ohm's law) builds on the same foundation. Volume IV
Chapter 5 develops the underlying physics.

\subsection{Mass}

The kilogram, since 20 May 2019, is realised through methods tied
to the fixed value of the Planck constant
$h = 6.626\,070\,15 \times 10^{-34}\,\si{\joule\second}$
(exact by definition) \cite{std:bipm-si2019,web:nist-kibble-balance}.
Two primary realisations carry the dissemination: the Kibble
balance and the X-ray crystal density (XRCD) method based on
ultrapure silicon-28 spheres.

A Kibble balance compares mechanical and electrical power in two
sequentially-operated modes. In weighing mode, the test mass on
one side of a balance is supported against gravity by a vertical
electromagnetic force $mg = BLI$, where $B$ is the magnetic field
in a radial-flux permanent magnet, $L$ is the effective coil
length, and $I$ is the current through the coil. In moving mode,
the test mass is removed and the coil is moved through the same
field at velocity $v$, generating an EMF $U = BLv$. Eliminating
the geometry-flux product $BL$ between the two modes gives
$mg = UI/v$, so the mass is determined by the electrical
quantities $U$ and $I$ (measurable through the Josephson and
quantum-Hall standards, hence through $h$ and $e$), the local
gravitational acceleration $g$ (measured with absolute gravimeters
to fractional uncertainty around $10^{-9}$), and the coil
velocity $v$ (measured with frequency-stabilised laser
interferometry to comparable precision). NIST's NIST-4 Kibble
balance at Gaithersburg, the NPL Kibble balance, and the BIPM
Kibble balance each realise the kilogram at total uncertainty in
the low parts-per-billion range; NIST-4's reported $1\sigma$
uncertainty is $13 \times 10^{-9}$ in fractional units, current
as of the 2017 measurements that supported the SI redefinition.

The XRCD method counts silicon atoms in a high-purity sphere. A
nearly-spherical single crystal of isotopically-enriched
silicon-28 (typically $99.99$\% $^{28}$Si, separated from
natural-abundance silicon by gas centrifugation at the Mendeleev
Institute in St Petersburg) is polished to sphericity better than
$30\,\si{\nano\meter}$ and characterised: its volume from optical
interferometry, its mass from comparison against a national mass
standard, and the lattice parameter $a$ from X-ray combined
optical interferometry give the atomic mass and hence Avogadro's
number $N_{A}$ via $m = (N_{A}/V_{\text{cell}})(V_{\text{sphere}}/
N)$ where $V_{\text{cell}} = a^{3}/8$ is the silicon unit cell
volume. The PTB silicon-sphere project achieved fractional
uncertainty $2 \times 10^{-8}$ on $N_{A}$ before the 2019
redefinition; the post-redefinition method runs in reverse,
delivering a kilogram realisation at the same uncertainty.

The dissemination chain. A Kibble-balance kilogram calibrates a
platinum-iridium working standard (a survivor of the artefact
era, now demoted from defining to working reference). The IPK and
its sister copies remain at the BIPM in Sevres and are
periodically compared. Working standards descend through:
\begin{itemize}[leftmargin=*]
  \item \engterm{Class E1} mass set (OIML R 111): assigned
    uncertainty $\pm 0.5\,\si{\milli\gram}$ on $1\,\si{\kilo\gram}$,
    density $7.95 \pm 0.10\,\si{\gram\per\centi\meter\cubed}$.
    Used by accredited laboratories.
  \item \engterm{Class E2}: $\pm 1.5\,\si{\milli\gram}$ on
    $1\,\si{\kilo\gram}$. Used by reference-laboratory calibrations.
  \item \engterm{Class F1}: $\pm 5\,\si{\milli\gram}$ on
    $1\,\si{\kilo\gram}$. Standard for high-precision industrial use.
  \item \engterm{Class F2, M1, M2, M3}: progressively coarser, with
    M3 at $\pm 500\,\si{\milli\gram}$ for general commercial use.
\end{itemize}
The working laboratory's analytical balance (a Mettler-Toledo
XPR205 at $0.01\,\si{\milli\gram}$ readability, repeatability
$0.02\,\si{\milli\gram}$ at $100\,\si{\gram}$ in a typical
metrology laboratory) is calibrated against a class E2 weight set
and inherits the chain's uncertainty.

The IPK's slow mass drift against its sister copies was the
practical argument for the 2019 redefinition: an artefact whose
own mass cannot be verified without circular reference is a poor
anchor for a unit. The documented drift, measured during the
three official BIPM periodic verifications (1899-1911, 1948-1953,
and 1988-1992) is approximately $50\,\si{\micro\gram}$ over the
$130$-year period, or $\approx 0.4\,\si{\micro\gram}$ per year on
average, with the direction of drift inconsistent across the
sister copies. The IPK case is registered as a standards-history
reference at
\repopath{docs/research/accidents/ipk-mass-drift-1889-2019.md}.

The post-redefinition dissemination chain has no measurable
discontinuity at the 2019 boundary: the redefinition's preparatory
work tied the new constants to the old artefact at every NMI
simultaneously to fractional uncertainty $\sim 10^{-8}$, so a
calibration certificate issued in 2018 and a certificate issued
in 2020 differ only in their cited constants, not in their
numerical content. The redefinition was deliberately engineered
for continuity \cite{std:bipm-si2019}.

\input{volumes/01-quantity/ch03-calibration-and-traceability/figures/fig-kibble-balance.tex}

\subsection{Length}

The metre is realised through frequency-stabilised laser
interferometry. The metre's definition pins it through the speed of
light $c$ and the second's definition: a metre is the length light
travels in vacuum during $1/c$ of a second, with $c = 299{,}792{,}458
\,\si{\meter\per\second}$ exact and the second realised through
cesium-133 hyperfine-transition fountain clocks. The CIPM's
recommended practical realisations of the metre
(\textit{mise en pratique}) list five working transitions and their
recommended frequencies: the $^{127}$I$_{2}$ R(127) 11-5 a$_{16}$
transition at $\nu = 473\,612\,353\,604\,\si{\kilo\hertz}$
(used by the iodine-stabilised HeNe at $633\,\si{\nano\meter}$), the
$^{40}$Ca-CH$_{4}$ system, the $^{85}$Rb 5S-5D two-photon
transition, the strontium-clock $^{87}$Sr $^{1}\!S_{0}{-}^{3}\!P_{0}$
optical transition at $429\,228\,004\,229\,873\,\si{\hertz}$, and the
$^{171}$Yb$^{+}$ ion $^{2}\!S_{1/2}{-}^{2}\!F_{7/2}$ octupole
transition. Each transition's realisation has a defined
fractional uncertainty: the iodine-stabilised HeNe reaches
$2 \times 10^{-11}$ at standard conditions, the strontium clock
reaches $2 \times 10^{-18}$ in the laboratory, and the optical
transitions are the basis for the projected second redefinition.

The dissemination chain for length is the longest of the working SI
quantities because length measurements span eleven orders of
magnitude in working practice. At the primary level, the NMI's
iodine-stabilised helium-neon laser at $632.991\,212\,58\,\si{\nano
\meter}$ is calibrated against the optical-frequency comb. At the
secondary level, gauge blocks are calibrated against the iodine-
laser interferometer using mechanical or optical comparison: the
gauge-block stack's length is measured by counting interference
fringes against the laser wavelength, with each fringe at
$\lambda/2 \approx 316.5\,\si{\nano\meter}$ providing the working
resolution. A class K (NMI-grade) gauge-block set carries a
calibration uncertainty around $20\,\si{\nano\meter}$ on the
$100\,\si{\milli\meter}$ block; the working laboratory's
class 0 set (purchased from Mitutoyo, Zeiss, or Mahr) inherits
$\pm 0.20\,\si{\micro\meter}$ uncertainty on the $100\,\si{\milli
\meter}$ block, a factor of ten larger. At the working level, a
hand-held micrometer with $25\,\si{\milli\meter}$ range and
$0.001\,\si{\milli\meter}$ readout carries a maximum permissible
error of $\pm 0.002\,\si{\milli\meter}$ per the
ISO 3611:2010 standard, two orders of magnitude beyond the gauge
block. Coordinate measuring machines (CMMs) such as the Zeiss
Prismo Ultra at $\pm (0.7 + L/600)\,\si{\micro\meter}$ where $L$
is in millimetres carry the dimensional discipline of the chapter's
discipline of precision measurement.

The silent failure mode in length metrology is thermal expansion.
Steel has a linear thermal expansion coefficient $\alpha \approx
11.5 \times 10^{-6}\,\si{\per\kelvin}$, so a $1\,\si{\meter}$ steel
gauge block at $1\,\si{\kelvin}$ above the $20\,\si{\degreeCelsius}$
standard reference temperature reads $11.5\,\si{\micro\meter}$
longer than its calibrated value. Aluminium ($\alpha = 23 \times
10^{-6}\,\si{\per\kelvin}$), titanium ($8.6 \times 10^{-6}$),
silicon ($2.6 \times 10^{-6}$), and Invar (a 36\% nickel-iron alloy
with $\alpha \approx 1.2 \times 10^{-6}$) span the working range.
Precision length measurements are conducted in temperature-
stabilised laboratories at the standard temperature, typically
controlled to within $\pm 0.1\,\si{\kelvin}$, with the gauge block's
temperature logged on the calibration certificate, the user
applying the temperature correction at the time of measurement, and
the working uncertainty including a contribution from the
combined coefficient-of-thermal-expansion uncertainty and the
measurement-temperature uncertainty.

\input{volumes/01-quantity/ch03-calibration-and-traceability/figures/fig-thermal-expansion.tex}

\textbf{Worked thermal correction.} A machine shop measures a
critical bore against a $100.000\,000\,\si{\milli\meter}$ steel gauge
block whose calibration certificate, issued at the
$20\,\si{\degreeCelsius}$ reference temperature, assigns a length of
$99.999\,920\,\si{\milli\meter}$ with expanded uncertainty
$U = 0.20\,\si{\micro\meter}$ at $k = 2$. The shop floor sits at
$23.5 \pm 0.3\,\si{\degreeCelsius}$; the block equilibrates to floor
temperature before use. The corrected length at use temperature is
\[
L(T) = L_{20}\,\bigl(1 + \alpha (T - 20)\bigr),
\]
with $\alpha = 11.5 \times 10^{-6}\,\si{\per\kelvin}$ for the
gauge-block steel. At $T = 23.5\,\si{\degreeCelsius}$, the fractional
expansion is $\alpha \Delta T = 11.5 \times 10^{-6} \times 3.5 =
4.025 \times 10^{-5}$, so
\[
L(23.5) = 99.999\,920 \times (1 + 4.025 \times 10^{-5})
= 99.999\,920 + 4.02\,\si{\micro\meter}
= 100.003\,94\,\si{\milli\meter}.
\]
The uncorrected reading is $4.0\,\si{\micro\meter}$ short of the
block's true length at floor temperature, twenty times the block's
own calibration uncertainty. A shop that skips the correction has
built a $4\,\si{\micro\meter}$ systematic error into every part
measured against the block.

The correction carries its own uncertainty. Two contributions matter.
The temperature is known to $\pm 0.3\,\si{\kelvin}$ ($u_{T} = 0.3 /
\sqrt{3} = 0.17\,\si{\kelvin}$ as a uniform bound), contributing
$L_{20}\,\alpha\,u_{T} = 100\,\si{\milli\meter} \times 11.5 \times
10^{-6}\,\si{\per\kelvin} \times 0.17\,\si{\kelvin} = 0.20\,\si{\micro
\meter}$. The expansion coefficient is itself uncertain by roughly
$\pm 1 \times 10^{-6}\,\si{\per\kelvin}$ for unspecified tool steel
($u_{\alpha} = 0.6 \times 10^{-6}\,\si{\per\kelvin}$), contributing
$L_{20}\,\Delta T\,u_{\alpha} = 100\,\si{\milli\meter} \times
3.5\,\si{\kelvin} \times 0.6 \times 10^{-6}\,\si{\per\kelvin} =
0.21\,\si{\micro\meter}$. Combined with the block's own
$u_{\text{cal}} = 0.10\,\si{\micro\meter}$ in quadrature,
\[
u_{c} = \sqrt{0.10^{2} + 0.20^{2} + 0.21^{2}}
= 0.31\,\si{\micro\meter},
\]
expanded to $U = 0.62\,\si{\micro\meter}$ at $k = 2$. Applying the
thermal correction shrinks the systematic error from
$4.0\,\si{\micro\meter}$ to zero and inflates the uncertainty from
$0.20$ to $0.62\,\si{\micro\meter}$: a trade the shop takes every
time, because a known $0.62\,\si{\micro\meter}$ uncertainty is worth
more than a hidden $4\,\si{\micro\meter}$ bias.

The correction's leading uncertainty is the expansion coefficient,
not the temperature, once the shop controls temperature to a few
tenths of a kelvin. A shop that wants to shrink the budget further
buys a gauge block of a material whose $\alpha$ is both small and
well-characterised: Figure~\ref{fig:vol01:ch03:thermal-expansion}
shows that an Invar block moves a tenth as far per kelvin as steel,
which is why long-baseline length standards and the reference frames
of coordinate measuring machines are built from low-expansion alloys
and, at the highest tier, from Zerodur or ULE glass-ceramic with
$\alpha$ near zero across the working range.

\subsection{Frequency}

The second is realised through cesium-133 fountain atomic clocks at
the major national metrology institutes. A fountain clock launches
a cloud of cesium atoms upward through a microwave cavity; the
atoms interact with the cavity's microwave field on the way up and
on the way down; the resonance condition for the cesium-133
hyperfine transition determines the cavity's frequency to fractional
uncertainty around $10^{-16}$. The defining constant is
$\Delta\nu_{\text{Cs}} = 9{,}192{,}631{,}770 \,\si{\hertz}$, exact
\cite{std:bipm-si2019}.

Frequency is the SI quantity realised at the lowest fractional
uncertainty of any base quantity; consequently, where engineering
quantities can be expressed as frequencies (timing intervals,
relative phases, ratios that reduce to frequency comparisons), they
can be measured at uncertainties orders of magnitude better than
direct measurements of position or amplitude. The trick of recasting
a measurement as a frequency comparison is one of the central tools
of precision metrology.

\textit{Primary realisations.} NIST-F2, operating at the Time and
Frequency Division in Boulder, Colorado, reports a Type B
fractional uncertainty of $1.1 \times 10^{-16}$ at its 2014
introduction, the lowest of any operating primary frequency
standard at the time; NIST-F2 supplements rather than replaces
NIST-F1, which continues operating with fractional uncertainty
$\sim 3 \times 10^{-16}$ \cite{web:ch03t3-nist-cesium-fountain-f2}.
NPL-CsF2 at Teddington reports Type B fractional uncertainty
$2.3 \times 10^{-16}$ \cite{web:ch03t3-npl-csf2-fountain}; PTB
CSF2 at Braunschweig reports $\sim 2 \times 10^{-16}$
\cite{web:ch03t3-ptb-csf2-fountain}. Together with PTB CSF1, SYRTE
(Paris) FO1/FO2, INRIM (Turin), KRISS (Korea), NMIJ-F1 (Japan),
and NIM5 (China), these standards anchor TAI through their reported
weights at the BIPM Time Department, where the international time
scale is computed from the weighted ensemble of approximately $450$
contributing clocks at $\sim 80$ laboratories worldwide
\cite{web:ch03t3-bipm-tai}.

\textit{Secondary representations.} A fountain clock is a research
instrument, not a working reference. The day-to-day dissemination of
frequency uses two secondary technologies, each occupying its own
performance-and-cost tier:

\begin{itemize}[leftmargin=*]
  \item \engterm{Active hydrogen masers}. The Microsemi MHM 2010 (a
    descendant of the Sigma-Tau and Symmetricom MHM lineages) is the
    canonical commercial hydrogen maser
    \cite{web:ch03t3-microsemi-mhm-2010}. Frequency stability
    $\sigma_{y}(\tau) = 1.5 \times 10^{-13}/\sqrt{\tau}$ for $1
    \leq \tau \leq 100 \,\si{\second}$, reaching a $10^{-15}$ floor
    at $\tau \sim 10^{4} \,\si{\second}$. Cost is in the range
    \$200{,}000 to \$400{,}000. Hydrogen masers serve as flywheel
    standards between fountain-clock evaluations at NMIs (the
    fountain provides absolute frequency, the maser provides
    short-term stability between fountain runs) and as the on-board
    oscillators for the ESA Galileo satellite payload.
  \item \engterm{Commercial cesium-beam clocks}. The Microsemi 5071A
    \cite{web:ch03t3-microsemi-5071a}, introduced as the HP/Agilent
    5071A in 1991, is the workhorse industrial cesium standard.
    Accuracy $\pm 1 \times 10^{-12}$ with a standard tube,
    $\pm 5 \times 10^{-13}$ with a high-performance tube; stability
    $1.2 \times 10^{-11}/\sqrt{\tau}$. The 5071A is the principal
    contribution to TAI from industrial timing laboratories and
    from telecom-grade Stratum 1 references. Cost is in the range
    \$70{,}000 to \$120{,}000.
  \item \engterm{Rubidium frequency standards}. The Spectratime
    LCR-900 (now Safran Electronics \& Defense)
    \cite{web:ch03t3-spectratime-lcr-900} and the Microsemi FE-5680A
    are representative rubidium oscillators at the \$2{,}000 to
    \$10{,}000 tier. Aging rate $\leq 5 \times 10^{-11}/\text{
    month}$, temperature coefficient $\leq 3 \times 10^{-10}$ over
    $-20$ to $+65 \,\si{\degreeCelsius}$. Rubidium clocks lack the
    cesium tube's absolute reference; they are tuned at the factory
    against a cesium reference and disciplined in service against
    GPS or against the laboratory's higher-tier standard.
\end{itemize}

\textit{Working oscillators.} Beneath the rubidium tier sit the
ovenised crystal oscillators that populate working instruments. The
Vectron MD-011 OCXO \cite{web:ch03t3-vectron-md-011} is
representative: frequency stability $\pm 5 \times 10^{-9}$ over
$-40$ to $+85 \,\si{\degreeCelsius}$, aging
$\leq 5 \times 10^{-10}/\text{day}$ after warmup, list price
approximately \$500--\$2{,}000. The Connor-Winfield CWX family
covers the same tier with slightly different stability-cost
trade-offs. An OCXO inside an instrument provides the instrument's
short-term frequency reference; the instrument's long-term
calibration comes from periodic comparison against a higher-tier
standard.

\begin{mastery}
Frequency standards are not characterised by a single standard
deviation, because their noise is not white. The stability figures
quoted above (the maser's $1.5 \times 10^{-13}/\sqrt{\tau}$, the
5071A's $1.2 \times 10^{-11}/\sqrt{\tau}$) are values of the
\engterm{Allan deviation} $\sigma_{y}(\tau)$, the standard tool for
frequency stability. The ordinary sample variance of the fractional
frequency does not converge for oscillators dominated by flicker or
random-walk noise: adding more data makes the sample variance grow
rather than settle. The Allan variance replaces the deviation from
the global mean with the difference between successive averages over
an averaging time $\tau$,
\[
\sigma_{y}^{2}(\tau) = \frac{1}{2}
\bigl\langle (\bar{y}_{k+1} - \bar{y}_{k})^{2} \bigr\rangle,
\]
where $\bar{y}_{k}$ is the mean fractional frequency over the $k$-th
interval of length $\tau$. The factor of one-half makes
$\sigma_{y}(\tau)$ equal to the classical standard deviation when the
noise is white, and the successive-difference structure makes the
statistic converge even when it is not.

Reading the log-log plot of $\sigma_{y}(\tau)$ against $\tau$ names
the dominant noise at each averaging time by its slope. White
frequency noise falls as $\tau^{-1/2}$; flicker frequency noise is
flat (the \engterm{flicker floor}); random-walk frequency noise rises
as $\tau^{+1/2}$. A working oscillator improves with averaging until
it hits its flicker floor, then degrades as random-walk and
environmental drift take over. The averaging time at the minimum is
where the oscillator is most stable and is the natural interval for a
frequency comparison against a higher-tier standard. A calibration
that averages far past the flicker floor wastes time and, worse,
imports the standard's own random-walk drift into the comparison. The
Allan deviation is why a frequency calibration certificate specifies
an averaging time alongside its uncertainty: without $\tau$, the
stability figure is unreadable.
\end{mastery}

\textit{GPS-disciplined oscillators.} A GPSDO uses GPS signals from
the satellite ensemble's onboard atomic clocks to discipline a
ground-based OCXO or rubidium oscillator, combining the OCXO's
short-term stability with the GPS ensemble's long-term accuracy.
The Trimble Thunderbolt (discontinued but widely deployed)
\cite{web:ch03t3-trimble-thunderbolt} and the Jackson Labs Fury
\cite{web:ch03t3-jackson-labs-fury} are canonical industrial
GPSDOs: holdover frequency stability $\pm 1.0 \times 10^{-10}$
over 24 hours after GPS loss, long-term frequency tracking to the
GPS ensemble $\pm 1.0 \times 10^{-12}$ over a 24-hour average.
A GPSDO at the \$1{,}000--\$5{,}000 tier provides frequency
accuracy comparable to a commercial cesium clock at $1/100$ of
the cost, conditional on a usable GPS sky view. The GPSDO's link
to the SI runs through the satellite ensemble's on-board atomic
clocks; the satellite clocks are themselves disciplined against
master control segment ground clocks at Schriever Space Force Base,
which are in turn referenced to USNO (the US Naval Observatory's
master clock) and through USNO into TAI. The GPS chain is the
forward-reference subject of Chapter 6 in this volume on time and
GPS.

\textit{Working instruments.} The Keysight 53230A universal
frequency counter \cite{web:ch03t3-keysight-53230a} is the
working-laboratory counter at the \$5{,}000--\$8{,}000 tier:
$12$-digit resolution per second of gate time, time-interval
resolution $20 \,\si{\pico\second}$ single-shot. The 53230A
inherits its frequency reference either internally (from its
OCXO) or externally (from a GPSDO, rubidium, or maser piped in
through its $10 \,\si{\mega\hertz}$ reference input). A
laboratory's choice of external reference determines the floor on
the counter's accuracy: an internal OCXO supports parts-per-billion
work, a GPSDO supports parts-per-trillion work, a maser supports
parts-per-femto work for short averaging times.

\textit{Household-grade end of the chain.} The mains frequency
monitor, available as a smartphone app or as a \$50 USB device,
displays the local mains frequency in real time. In North America,
the mains is held within $\pm 0.02 \,\si{\hertz}$ of $60 \,\si{
\hertz}$ on average over a 24-hour period by the
ISO/RTO frequency-regulation services; in Europe, within
$\pm 0.05 \,\si{\hertz}$ of $50 \,\si{\hertz}$. The mains
frequency derives from generator-set rotational speeds that are
themselves controlled against grid-operator clocks traceable to
UTC. The household chain runs: the user's smartphone app shows
mains frequency, the mains frequency tracks the grid operator's
control clock, the control clock tracks UTC at the regional time
laboratory, UTC is the BIPM-published average of the contributing
clocks, and the contributing clocks are evaluated against the
fountain clocks. The chain has perhaps eight visible links from
the user's app to the cesium fountain; the user's confidence in
the displayed frequency depends on every link.

Frequency is the SI quantity for which the gap between primary
realisation ($10^{-16}$) and household working instrument
($10^{-7}$) spans nine orders of magnitude. The intermediate tiers
of masers, cesium beams, rubidium clocks, OCXOs, and GPSDOs each
fill a specific decade with a specific cost-and-performance
trade-off; the chapter's chain discipline applies at every level.

\subsection{Temperature}

The kelvin's realisation is the most operationally diverse of the
seven base units. Since 2019 the kelvin is defined through the
Boltzmann constant $k_{\text{B}} = 1.380\,649 \times 10^{-23}
\,\si{\joule\per\kelvin}$ (exact by definition)
\cite{std:bipm-si2019}. Primary realisations work across the full
operational range:

\engterm{Acoustic gas thermometry (AGT)}. The speed of sound in a
monatomic ideal gas at low pressure is $c = \sqrt{\gamma R T/M}$
where $\gamma = 5/3$, $M$ is the molar mass, and $R = N_{A} k_{B}$
is the molar gas constant. Measuring the acoustic resonance
frequencies of a high-purity argon-filled spherical cavity at
known pressure extracts $T$ to fractional uncertainty
$\sim 10^{-7}$ at room temperature. NPL's AGT realisation
contributed to the 2017 Boltzmann-constant value that fixed
$k_{B}$ at the 2019 redefinition. The working range is
$\sim 1$-$1235\,\si{\kelvin}$.

\engterm{Dielectric-constant gas thermometry (DCGT)}. The
permittivity of helium at known pressure is $\epsilon - 1 =
A_{\epsilon} \rho + B_{\epsilon} \rho^{2} + \cdots$, where
$A_{\epsilon}$ is theoretically known from ab initio quantum
chemistry of helium ($A_{\epsilon} = 0.517\,254\,06\,\si{\centi
\meter\cubed\per\mol}$). Measuring the helium-filled capacitor's
capacitance at known pressure extracts the density and hence the
temperature. PTB's DCGT realisation reaches fractional
uncertainty $\sim 4 \times 10^{-6}$ over the $5$-$300\,\si{\kelvin}$
range.

\engterm{Radiation thermometry}. Planck's spectral-radiance law
$L(\lambda, T) = (2 h c^{2}/\lambda^{5})/(\exp(hc/\lambda k_{B} T)
- 1)$ relates the radiance of a blackbody to its temperature.
Fitting the spectrum of a high-temperature blackbody cavity
extracts $T$ from the radiance at multiple wavelengths.
Filter-radiometer realisations at NIST and PTB reach fractional
uncertainty $\sim 10^{-4}$ at $T = 2400\,\si{\kelvin}$, with the
working range extending to $\sim 3000\,\si{\kelvin}$ for tungsten
strip lamps and higher for plasma-based radiance standards.

For working laboratories, the kelvin is disseminated through the
\textit{International Temperature Scale of 1990} (ITS-90)
\cite{web:bipm-cct-its90}. ITS-90 is a practical scale defined by
fifteen primary fixed points and ten standard interpolation
equations covering the range $0.65$-$1357.77\,\si{\kelvin}$
($-272.50$ to $+1084.62\,\si{\degreeCelsius}$). The fixed points
include the triple points of equilibrium hydrogen
($13.8033\,\si{\kelvin}$), neon ($24.5561$), oxygen ($54.3584$),
argon ($83.8058$), and water ($273.16\,\si{\kelvin}$, exact in
ITS-90), the melting point of gallium ($302.9146$), and the
freezing points of indium ($429.7485$), tin ($505.078$), zinc
($692.677$), aluminium ($933.473$), silver ($1234.93$), gold
($1337.33$), and copper ($1357.77$). The standard interpolation
instruments are: the constant-volume gas thermometer in the
cryogenic range, the platinum resistance thermometer (SPRT) in
the $13.8033$ to $1234.93\,\si{\kelvin}$ range using polynomial
deviation functions specific to each subrange, and the radiation
thermometer per Planck's law above $1234.93\,\si{\kelvin}$.

ITS-90 approximates thermodynamic temperature over its defined
range. The largest known ITS-90-to-thermodynamic deviation is
$\sim 10\,\si{\milli\kelvin}$ in the cryogenic range and
$\sim 50\,\si{\milli\kelvin}$ at the higher fixed points; CCT
working groups continue to refine the deviation values
\cite{web:bipm-cct-its90}. Working thermocouples (type R: Pt-13Rh
versus Pt; type K: Ni-Cr versus Ni-Al; type N), platinum
resistance thermometers (industrial Pt100 to $\pm 0.15\,\si{\kelvin}$
class B, $\pm 0.10$ class A), and radiation pyrometers are
calibrated against ITS-90 fixed points or against transfer
instruments traceable to them. A standard platinum resistance
thermometer (SPRT, Pt100 long-stem at $25.5\,\si{\ohm}$ resistance
ratio reference) is the working metrology reference; the worked
case in the next subsection walks through an SPRT ITS-90
calibration end to end.

A working temperature engineer sees instrument families separated
by orders of magnitude in cost and capability. A consumer-grade
type-K thermocouple with handheld reader (Fluke 51-II at
$\pm 1.1\,\si{\kelvin}$ accuracy or $\pm 0.4\%$ of reading,
whichever is greater) costs around \$200; an industrial Pt100
with four-wire transmitter (Yokogawa YTA710 at $\pm 0.07\,\si{
\kelvin}$ over the $-200$ to $+850\,\si{\degreeCelsius}$ range)
costs around \$3{,}000; an SPRT for accredited calibration
(Hart Scientific 5681 long-stem at $\pm 0.005\,\si{\kelvin}$
$k = 2$ over the working subrange) costs around \$15{,}000. The
chapter's discipline applies at every cost level: the instrument's
uncertainty must match the application's tolerance, and the
calibration chain must run back to ITS-90 fixed points with
documented uncertainty at every link.

\subsection{Worked case: ITS-90 calibration of a standard
platinum resistance thermometer}

A working calibration laboratory receives a long-stem standard
platinum resistance thermometer (SPRT) for its annual ITS-90
calibration. The instrument is a $25.5\,\si{\ohm}$ SPRT
manufactured to ITS-90 specifications, used by the customer as a
working reference for industrial RTD calibration in the
$0$-$420 \,\si{\degreeCelsius}$ range. The worked case walks
through the calibration end-to-end: fixed-point measurement,
deviation-function fitting, uncertainty budget, and reported
result.

\textbf{Setup and fixed points.} The SPRT's calibration covers the
ITS-90 sub-range from the triple point of water (TPW) to the
freezing point of zinc (FPZn). ITS-90 specifies four fixed points
for this range \cite{web:bipm-cct-its90}: the TPW
($273.16 \,\si{\kelvin}$, defined exact), the freezing point of
tin (FPSn, $231.928 \,\si{\degreeCelsius}$), the freezing point of
zinc (FPZn, $419.527 \,\si{\degreeCelsius}$), and the melting point
of gallium (MPGa, $29.7646 \,\si{\degreeCelsius}$). Each fixed
point is realised in a sealed cell maintained by the laboratory;
the SPRT is immersed in each cell sequentially and its resistance
is measured at the fixed-point plateau.

\textbf{Step 1: TPW measurement.} The SPRT is immersed in the TPW
cell. The cell is prepared by partially freezing distilled water
inside an outer cylindrical jacket; the SPRT's sensing element sits
at the ice-water-vapour triple-point boundary. The resistance is
measured after a thermal-equilibration period of $\geq 30$
minutes. The measurement: $R(TPW) = R_{0} = 25.4983 \,\si{\ohm}$.
This becomes the SPRT's defining $R_{0}$ for all subsequent
calculations.

\textbf{Step 2: FPSn measurement.} The SPRT is moved to the tin
freezing-point cell, immersed at the plateau, and read. The cell
is initiated by raising the cell temperature to slightly above the
melting point, then allowing controlled cooling; the freezing
plateau is identified by the rate-of-change reaching $\leq
0.0001 \,\si{\degreeCelsius}/\text{min}$. Measurement: $R(\text{
FPSn}) = 48.6125 \,\si{\ohm}$. The resistance ratio $W(\text{FPSn})
= R(\text{FPSn})/R_{0} = 1.906\,89$.

\textbf{Step 3: FPZn measurement.} Same protocol at the zinc cell.
Measurement: $R(\text{FPZn}) = 67.8434 \,\si{\ohm}$. Ratio
$W(\text{FPZn}) = 2.661\,11$.

\textbf{Step 4: MPGa measurement.} Same protocol at the gallium
cell. Measurement: $R(\text{MPGa}) = 28.4761 \,\si{\ohm}$. Ratio
$W(\text{MPGa}) = 1.116\,80$.

\textbf{Step 5: deviation-function fitting.} ITS-90 prescribes a
reference function $W_{r}(T)$ (a tabulated polynomial in
$\ln[T/273.16]$) plus a sub-range-specific deviation function
$\Delta W(T)$. For the $0$-$\text{FPZn}$ sub-range, the deviation
function is
\[
\Delta W(T) = a [W(T) - 1] + b [W(T) - 1]^{2} + c [W(T) - 1]^{3},
\]
with the two-coefficient sub-range to TPW-MPGa-FPSn-FPZn requiring
solving four equations:
\begin{align*}
W(\text{MPGa}) - W_{r}(\text{MPGa}) &= \Delta W(\text{MPGa}), \\
W(\text{FPSn}) - W_{r}(\text{FPSn}) &= \Delta W(\text{FPSn}), \\
W(\text{FPZn}) - W_{r}(\text{FPZn}) &= \Delta W(\text{FPZn}).
\end{align*}
This is a linear system in three unknowns $a, b, c$. The
laboratory's computer fits the coefficients to typically
$a \approx -1.1 \times 10^{-5}$, $b \approx -3.0 \times 10^{-6}$,
$c \approx 0$. The fitted deviation function permits computing
$T$ from any measured $W$ in the sub-range.

\textbf{Step 6: uncertainty budget.} The budget combines Type A
and Type B contributions. Representative values for a working
SPRT calibration:
\begin{itemize}
  \item Fixed-point realisation (TPW): $u = 0.10 \,\si{\milli
  \kelvin}$ (Type B, from cell design and impurity)
  \item Fixed-point realisation (FPSn): $u = 0.40 \,\si{\milli
  \kelvin}$ (Type B)
  \item Fixed-point realisation (FPZn): $u = 0.60 \,\si{\milli
  \kelvin}$ (Type B)
  \item Fixed-point realisation (MPGa): $u = 0.20 \,\si{\milli
  \kelvin}$ (Type B)
  \item Resistance bridge uncertainty: $u = 0.05 \,\si{\milli
  \kelvin}$ at TPW, scaling with $W$ at other points (Type B,
  from bridge calibration certificate)
  \item Self-heating correction: $u = 0.10 \,\si{\milli\kelvin}$
  (Type B, from the $I$-squared extrapolation to zero current)
  \item Hydrostatic head correction: $u = 0.05 \,\si{\milli
  \kelvin}$ (Type B)
  \item Stem conduction: $u = 0.10 \,\si{\milli\kelvin}$ (Type B)
  \item Repeatability across the calibration session:
  $u = 0.30 \,\si{\milli\kelvin}$ (Type A, from $n = 5$ plateau
  re-readings)
\end{itemize}
At the FPZn point, the contributions combine in quadrature:
\[
u(T)_{\text{FPZn}} = \sqrt{0.6^{2} + 0.4^{2} + 0.2^{2} + 0.1^{2}
+ 0.05^{2} + 0.1^{2} + 0.3^{2}} \approx 0.83 \,\si{\milli\kelvin}.
\]
At intermediate temperatures, the deviation-function fit
introduces an additional uncertainty of $\sim 0.5 \,\si{\milli
\kelvin}$ (interpolation residual). Combined uncertainty at any
intermediate $T$: $u(T) \approx 1.0 \,\si{\milli\kelvin}$.

\textbf{Step 7: reported result.} The certificate reports the
fixed-point resistances, the fitted deviation-function coefficients,
the combined uncertainty as a function of temperature, and the
coverage factor $k = 2$. Expanded uncertainty at the customer's
working range:
$U(T) = 2 \times 1.0 = 2.0 \,\si{\milli\kelvin}$.

\textbf{Sanity check.} Industrial RTDs operating in the $0$-$420
\,\si{\degreeCelsius}$ range commonly carry calibration
uncertainties of $\pm 30$-$100 \,\si{\milli\kelvin}$. The SPRT
calibration's $\pm 2 \,\si{\milli\kelvin}$ is $\sim 25\times$
smaller than the working RTD's, providing the required ratio of
working-to-reference uncertainty. The SPRT is fit for purpose as a
working reference standard.

\textbf{Implication for the chain.} Below the SPRT, customer RTDs
are calibrated against this SPRT in a comparison furnace; their
uncertainty inherits the SPRT's $2 \,\si{\milli\kelvin}$ plus the
furnace gradient and RTD-vs-SPRT comparison uncertainties,
typically yielding $\pm 20$-$50 \,\si{\milli\kelvin}$ at the
customer level. Above the SPRT, the SPRT itself was calibrated by
a national metrology institute against ITS-90 cells whose
realisations trace through CIPM key comparisons; the laboratory
imports the NMI's own $\sim 0.3 \,\si{\milli\kelvin}$ FPZn
uncertainty as the input to its budget.

\textbf{What the case teaches.} Calibration is a chain of
sub-budgets. Each step inherits the uncertainty of the level
above and adds its own contributions. The ITS-90 framework
specifies the protocol so each step is reproducible across
laboratories; the CIPM MRA \cite{std:bipm-cipm-mra} ensures the
NMI-level cells produce comparable realisations. A reader's own
calibration certificate, examined in this light, should display
the same layered structure: fixed points or reference values used,
deviation or correction function, contribution table, combined
$u$, expanded $U$, coverage factor.

\subsection{Pressure and force}

Pressure is realised mechanically. The primary standard is the
\engterm{dead-weight gauge} (also called the pressure balance or
piston gauge): a precisely-machined piston of effective area $A_{e}$
is loaded with calibrated masses $m_{i}$ under the local
gravitational acceleration $g_{\text{loc}}$, and the resulting
pressure
\[
p = \frac{(m_{\text{piston}} + \sum m_{i})\, g_{\text{loc}}\,
(1 - \rho_{\text{air}}/\rho_{\text{mass}})}
{A_{e}\,(1 + (\alpha_{p} + \alpha_{c})\,(T - T_{0}))}
+ p_{\text{atm}}
\]
is computed from first principles. The numerator carries the
air-buoyancy correction on the masses (with $\rho_{\text{air}}$ the
laboratory air density and $\rho_{\text{mass}}$ the stainless-steel
mass density $7.95\,\si{\gram\per\centi\meter\cubed}$). The
denominator carries the thermal-expansion correction with
$\alpha_{p}$ and $\alpha_{c}$ the linear-expansion coefficients of
piston and cylinder, typically tungsten carbide
($4.5 \times 10^{-6}\,\si{\per\kelvin}$) or hardened steel
($11 \times 10^{-6}\,\si{\per\kelvin}$) \cite{std:ch03t3b-oiml-r110-2007}.
The piston rotates during measurement at $5$-$30\,\si{rpm}$ to
average out the gap between piston and cylinder, and the gauge is
operated only when the piston is freely floating, with the load
column visibly suspended on the thin oil or gas film. The
effective area $A_{e}$ is determined by cross-floating against a
higher-level dead-weight gauge per EURAMET cg-3
\cite{std:ch03t3b-euramet-cg-3}, or by dimensional measurement
combined with elastic-distortion modelling, to fractional
uncertainty $\sim 10^{-5}$.

\textbf{Primary tier.} National metrology institutes operate
hydraulic and pneumatic dead-weight gauges across the
$1\,\si{\pascal}$ to $1\,\si{\giga\pascal}$ range. NIST's pressure
and vacuum laboratory \cite{web:ch03t3b-nist-pressure} maintains
two oil-operated piston-gauge primary standards, the NIST PG-7607
at $1\,\si{\mega\pascal}$ and a $17\,\si{\mega\pascal}$ system, with
combined-standard uncertainty in the low parts-per-million range.
DH-Budenberg of Manchester, UK, supplies the model 580-series
hydraulic dead-weight gauges to NMIs and secondary calibration
laboratories at accuracies of $0.005\%$ to $0.015\%$ of reading
across $1\,\si{\mega\pascal}$ to $1000\,\si{\mega\pascal}$
\cite{web:ch03t3b-dh-budenberg}; Ruska (now Fluke Calibration) builds
the gas-bearing Ruska 2465A pneumatic piston gauge for the
$7\,\si{\kilo\pascal}$ to $7\,\si{\mega\pascal}$ range at $0.0035\%$
of reading \cite{web:ch03t3b-ruska-2465}. The dead-weight gauge's
traceability decomposes into the mass calibration of the loading
weights (to the kilogram chain), the local-gravity measurement
(to absolute gravimeters, fractional uncertainty $10^{-8}$), and
the effective-area determination (to length, through the silicon-
sphere or laser-interferometer chains). A pressure measurement
is, at root, a mass and length measurement under known
gravitational acceleration.

\textbf{Secondary tier.} Accredited calibration laboratories
maintain piston gauges as transfer standards. The Fluke 2700G
portable pressure calibrator carries $\pm 0.02\%$ of reading
across $0$-$70\,\si{\mega\pascal}$ \cite{web:ch03t3b-fluke-2700g}.
The WIKA CPB5000 hydraulic and pneumatic variants reach $0.005\%$
of reading \cite{web:ch03t3b-wika-cpb5000}, sitting between the
NMI primary and the working transmitter. Digital reference
manometers form a parallel secondary tier when the laboratory
does not require the dead-weight gauge's full uncertainty. The
Mensor CPC6050 modular pressure controller achieves $\pm 0.01\%$
of full scale across a configurable range
\cite{web:ch03t3b-mensor-cpc6050} and offers automated
transmitter-calibration sequences without manual mass loading.

\textbf{Working tier.} Industrial pressure transmitters dominate
the working laboratory. The Rosemount 3051 with the high-end
specification reaches $\pm 0.025\%$ of upper range limit (URL)
reference accuracy, with the standard specification at $0.075\%$
URL and the basic specification at $0.10\%$ URL
\cite{web:ch03t3b-rosemount-3051}; Honeywell's ST 800 SmartLine
line ranges from $\pm 0.0375\%$ URL (Advanced) to $\pm 0.1\%$ URL
(Basic) \cite{web:ch03t3b-honeywell-st800}; Endress+Hauser's
Cerabar PMP71 specifies $\pm 0.025\%$ of span at the Platinum
accuracy class and $\pm 0.075\%$ at the Standard class
\cite{web:ch03t3b-eh-cerabar}. The 0.025\% to 0.5\% accuracy span
across these working instruments reflects the trade-off between
the transmitter's silicon-strain-gauge sensor and the user's
budget; a $0.025\%$ URL transmitter costs roughly $4\times$ what
the $0.5\%$ URL transmitter costs from the same manufacturer.

\textbf{Force.} Force is realised through dead-weight machines:
calibrated masses act under local gravity to produce a known
force on a load cell or proving ring under test. NIST's $1\,\si{
\mega\newton}$ deadweight machine and the PTB $1\,\si{\mega
\newton}$ machine are the upper end of the chain. Working
laboratories use force transducers (HBM C2A series at $\pm 0.05\%$
class, Interface 1000-series at $\pm 0.04\%$ class) calibrated
against secondary deadweight machines or against an NMI-traceable
reference transducer per ISO 376. The chain runs back to mass and
length and inherits the kilogram's $\sim 10^{-8}$ realisation
uncertainty plus the local-gravity uncertainty plus the proving
machine's mechanical contributions.

\textbf{Forward link.} The pressure chain reappears in
Volume I Chapter 7 (materials), where the elastic-distortion
correction in the piston-cylinder assembly is computed from the
materials' Young's modulus and Poisson ratio. The same chain
reappears in Volume V (continuum mechanics and fluids), where
the pressure transmitter's installation conditions, head
correction, and orientation effects determine whether the
reported value corresponds to the process variable of interest.

\textbf{Worked cross-float comparison.} A working pressure
laboratory operates two piston gauges: a $0$-$10\,\si{\mega
\pascal}$ hydraulic reference (WIKA CPB5000 with $A_{e,\text{ref}}
= 8.398\,512 \times 10^{-5}\,\si{\meter\squared}$ from its
calibration certificate) and a $0$-$10\,\si{\mega\pascal}$
unit-under-test piston gauge whose effective area is to be
determined. The cross-float comparison places both gauges on a
common pressure manifold, loads each piston with calibrated mass
sets that produce nominally equal pressures, and adjusts the load
on the unit-under-test until the two pistons stop drifting
relative to one another (the "float balance" condition). At
balance, the two gauges report the same pressure:
\[
\frac{(m_{\text{ref}} + m_{p,\text{ref}})\, g_{\text{loc}}\,
B_{\text{ref}}}{A_{e,\text{ref}}\,(1 + \beta_{\text{ref}} \Delta T)}
= \frac{(m_{\text{uut}} + m_{p,\text{uut}})\, g_{\text{loc}}\,
B_{\text{uut}}}{A_{e,\text{uut}}\,(1 + \beta_{\text{uut}} \Delta T)},
\]
where $B = (1 - \rho_{\text{air}}/\rho_{\text{mass}})$ is the
air-buoyancy correction and $\beta = \alpha_{p} + \alpha_{c}$ is
the sum of piston and cylinder linear-expansion coefficients
\cite{std:ch03t3b-euramet-cg-3}. The local gravity
$g_{\text{loc}} = 9.801\,234\,\si{\meter\per\second\squared}$
(measured by absolute gravimeter to fractional uncertainty $10^{-8}$)
cancels from the comparison; the air-buoyancy corrections
$B_{\text{ref}} \approx 0.999\,851$ and $B_{\text{uut}} \approx
0.999\,851$ effectively cancel when both piston-mass assemblies
use the same stainless-steel density, leaving a near-pure
geometric comparison.

Working through the balance condition at $p = 5\,\si{\mega\pascal}$,
$T = 23.00 \pm 0.02\,\si{\degreeCelsius}$:
\begin{itemize}[leftmargin=*]
  \item Reference loaded mass: $m_{\text{ref}} + m_{p,\text{ref}}
    = 42.836\,425\,\si{\kilo\gram}$ (from the calibrated mass
    set, traceable to Class E2).
  \item UUT loaded mass at balance: $m_{\text{uut}} +
    m_{p,\text{uut}} = 43.211\,852\,\si{\kilo\gram}$ (the
    operator added masses in $1\,\si{\gram}$ increments until
    the piston-drift rate fell below $0.1\,\si{\milli\meter}/
    \text{min}$).
  \item Ratio of loaded masses: $43.211\,852 / 42.836\,425
    = 1.008\,766$.
\end{itemize}
With both gauges using the same tungsten-carbide piston-cylinder
material ($\beta = 9.0 \times 10^{-6}\,\si{\per\kelvin}$) and the
same $T$, the temperature correction cancels exactly. The
effective-area ratio is therefore
\[
\frac{A_{e,\text{uut}}}{A_{e,\text{ref}}} = 1.008\,766,
\quad A_{e,\text{uut}} = 1.008\,766 \times 8.398\,512 \times
10^{-5}\,\si{\meter\squared} = 8.471\,150 \times 10^{-5}\,\si{
\meter\squared}.
\]
The uncertainty of this determination combines: the reference
gauge's $\pm 5 \times 10^{-6}$ fractional area uncertainty (from
its certificate), the mass-set uncertainty (Class E2: $\pm 1.5
\,\si{\milli\gram}$ on a $1\,\si{\kilo\gram}$ piece, giving
$\pm 4 \times 10^{-6}$ at the full $43\,\si{\kilo\gram}$ load
through quadrature), the balance-resolution uncertainty
($\pm 0.5\,\si{\gram}$ at the minimum-increment step, giving
$\pm 1.2 \times 10^{-5}$ fractional), and the temperature-match
uncertainty ($\pm 0.02\,\si{\kelvin}$ at $\beta = 9 \times
10^{-6}\,\si{\per\kelvin}$ gives $\pm 1.8 \times 10^{-7}$, a
negligible contribution because both gauges share the same
temperature). Combining in quadrature:
\[
\frac{u(A_{e,\text{uut}})}{A_{e,\text{uut}}} = \sqrt{
(5 \times 10^{-6})^{2} + (4 \times 10^{-6})^{2} +
(1.2 \times 10^{-5})^{2}}
\approx 1.4 \times 10^{-5}.
\]
Expanded at $k = 2$: $\pm 2.8 \times 10^{-5}$, or
$\pm 2.4\,\si{\milli\meter\squared}/\si{\meter\squared}$ on the
effective area. The cross-float determination of the UUT's
effective area is now traceable through the reference gauge,
through the CPB5000's NMI calibration, to NIST's PG-7607 piston
gauge, to NIST's master mass set, to NIST-4 Kibble balance.
Each step in the chain is documented; each contributes a
quantified piece to the UUT's overall uncertainty.

The cross-float comparison is the working pressure laboratory's
analogue of the SPRT comparison furnace and the Zener-DMM
voltage comparison: a controlled physical comparison between two
artefacts with carefully characterised auxiliary corrections,
performed against a higher-tier reference whose own uncertainty
is documented.

\textbf{Historical sketch.} The dead-weight gauge predates the SI
by more than a century. Joseph Bramah's hydraulic press patent of
1795 contains the principle: pressure is transmitted equally
through a confined fluid, so a calibrated weight on a small
piston balances a much larger load on a large piston. Bramah's
contemporaries used the apparatus as a working tool, not as a
metrological reference; the conceptual shift to traceable
pressure standard occurred in the late nineteenth century. By
the 1880s, the German Physikalisch-Technische Reichsanstalt
(PTR, established 1887, the institutional ancestor of PTB
founded 1953) operated dead-weight gauges as transfer
references for the boiler-pressure industry. The 1901 founding
of the US National Bureau of Standards
\cite{hist:ch03t3-nbs-act-1901} placed pressure metrology under
the same national-laboratory roof as electrical and length
metrology. NIST's pressure and vacuum laboratory traces its
piston-gauge realisations through a continuous line of
intercomparisons against PTB and NPL since the 1920s
\cite{web:ch03t3b-nist-pressure}, with the CCM key comparison
CCM.P-K series providing the post-1990 international consistency
check. The accreditation framework (ISO/IEC 17025) entered
pressure metrology in the late 1990s and is now the universal
working standard for commercial pressure-calibration
laboratories.

\textbf{Forces of the chain.} A pressure laboratory's working
discipline at the 2026 frontier includes one practical
complication: real pressure gauges are not point sensors. The
hydraulic dead-weight gauge has a column of oil of finite height
between the piston and the test port, and the column's
contribution to the measured pressure ($\rho g h$ where $\rho$
is the oil density, $g$ the local gravity, $h$ the column
height) must be added or subtracted depending on the geometry.
A $1\,\si{\meter}$ column of typical calibration oil
($\rho \approx 870\,\si{\kilo\gram\per\meter\cubed}$) contributes
$8.5\,\si{\kilo\pascal}$ to the measured pressure, a contribution
that dwarfs the $\pm 1\,\si{\kilo\pascal}$ calibration
uncertainty derived in the worked budget. The head correction is
the dominant systematic contribution at low-pressure points and
the leading uncertainty contribution when the column height is
not precisely characterised. A laboratory whose calibration
certificates do not state the column height, the oil density,
and the orientation has elided the largest single contribution
to the budget.

\subsection{Worked chain budget: a $10\,\si{\volt}$ DMM calibration}

The electrical analogue of the SPRT case demonstrates the same
chain discipline in a different physical domain. An accredited
calibration laboratory calibrates a customer's Keysight 3458A
8.5-digit bench DMM at the $10\,\si{\volt}$ DC point against the
laboratory's transfer-grade Zener voltage reference, which was
itself calibrated against a National Metrology Institute's
Josephson voltage standard six months earlier. The budget
follows the same Type A and Type B contribution structure as the
SPRT case; the numerical values reflect typical accredited-
laboratory practice at the parts-per-million level.

\begin{center}
\small
\begin{tabular}{lccccc}
\toprule
Source & Type & Distr. & $u(x_{i})$ ($\si{\micro\volt}$) & $c_{i}$ & Contribution \\
\midrule
NMI Josephson reference & B & Gaussian & $0.30$ & $1$ & $0.30\,\si{\micro\volt}$ \\
Zener drift since cal & B & Uniform & $0.40$ & $1$ & $0.40\,\si{\micro\volt}$ \\
Zener-DMM comparison repeatability & A & Gaussian & $0.35$ & $1$ & $0.35\,\si{\micro\volt}$ \\
Thermal EMFs in cables & B & Uniform & $0.25$ & $1$ & $0.25\,\si{\micro\volt}$ \\
Temperature stability ($\pm 0.5\,\si{\kelvin}$) & B & Uniform & $0.18$ & $1$ & $0.18\,\si{\micro\volt}$ \\
DMM internal noise floor & A & Gaussian & $0.12$ & $1$ & $0.12\,\si{\micro\volt}$ \\
Mains-frequency-induced pickup & B & Uniform & $0.08$ & $1$ & $0.08\,\si{\micro\volt}$ \\
Resistor self-heating in switching & B & Uniform & $0.05$ & $1$ & $0.05\,\si{\micro\volt}$ \\
\midrule
Combined $u_{c}(V)$ & & & & & $0.74\,\si{\micro\volt}$ \\
Expanded $U(V)$ at $k = 2$ & & & & & $1.5\,\si{\micro\volt}$ \\
\bottomrule
\end{tabular}
\end{center}

The combined standard uncertainty is
$u_{c} = \sqrt{0.30^{2} + 0.40^{2} + 0.35^{2} + 0.25^{2} +
0.18^{2} + 0.12^{2} + 0.08^{2} + 0.05^{2}} = \sqrt{0.090 + 0.160
+ 0.123 + 0.063 + 0.032 + 0.014 + 0.006 + 0.003} = \sqrt{0.491}
\approx 0.70\,\si{\micro\volt}$ (the slight discrepancy from the
$0.74\,\si{\micro\volt}$ tabulated above reflects intermediate
rounding in the table). The expanded uncertainty at $k = 2$ is
$U(V) \approx 1.5\,\si{\micro\volt}$, or $0.15$ parts per million
of the $10\,\si{\volt}$ measurement.

The certificate would report: ``At $10\,\si{\volt}$ DC the
instrument reads $V_{\text{ref}} + (0.4 \pm 1.5)\,\si{\micro
\volt}$ at $k = 2$, valid for a $\pm 1\,\si{\kelvin}$
environmental range around the calibration temperature, with a
recommended recalibration interval of one year.''

The Keysight 3458A's manufacturer-stated $24$-hour accuracy
specification at $10\,\si{\volt}$ DC is $\pm 0.5\,\si{ppm}$ of
reading plus $\pm 0.05\,\si{ppm}$ of range, equivalent to
$\pm 5.5\,\si{\micro\volt}$ at $k = 2$ (assuming the manufacturer
spec corresponds to $k = 2$ Gaussian coverage); the laboratory's
calibration uncertainty of $\pm 1.5\,\si{\micro\volt}$ is $\sim
4\times$ tighter than the manufacturer's specification. The
$4{:}1$ ratio is the working metrological convention for
reference-to-tolerance matching: the calibration uncertainty
should be at least four times tighter than the instrument's
specification for the calibration to be the dominant source of
working knowledge about the instrument's actual error.

\textbf{Implication for the chain.} Below the laboratory's
Zener-and-DMM working comparison, customer measurements with
this DMM at $10\,\si{\volt}$ DC inherit the $\pm 1.5\,\si{\micro
\volt}$ calibration uncertainty plus contributions from the
customer's own measurement chain (probe leads, thermal EMFs,
analog-to-digital conversion settling time, autozero stability).
A typical customer measurement carries a combined uncertainty of
$\pm 5$ to $\pm 10\,\si{\micro\volt}$ at $k = 2$ for the same
$10\,\si{\volt}$ point.

Above the laboratory's working comparison, the Zener reference
inherits the NMI's Josephson-derived uncertainty of $\pm 0.3\,\si{
\micro\volt}$ at $k = 1$ from the original NMI calibration
certificate. The NMI's uncertainty derives from the BIPM-
coordinated CIPM key comparisons (CCEM-K11, the routine
intercomparison of $10\,\si{\volt}$ Josephson standards across
NMIs) and from the local Josephson array's documented
uncertainty budget at the cryogenic operating point.

\subsection{Worked chain budget: a pressure-transmitter
calibration}

The pressure analogue of the DMM case walks the same chain
discipline through a mechanical realisation. A working petrochemical
plant operates a Rosemount 3051CG4A pressure transmitter, $0$ to
$10\,\si{\mega\pascal}$ range with the high-end $\pm 0.025\%$ URL
reference accuracy \cite{web:ch03t3b-rosemount-3051}. The plant's
in-house calibration laboratory calibrates the transmitter at the
$5\,\si{\mega\pascal}$ midpoint against a Fluke 2700G portable
piston-gauge reference \cite{web:ch03t3b-fluke-2700g}, which was
itself calibrated against the plant's WIKA CPB5000 hydraulic
pressure balance \cite{web:ch03t3b-wika-cpb5000} six months
earlier, which had been cross-floated against an NMI primary
dead-weight gauge eighteen months earlier. The chain has three
visible links above the transmitter.

The budget at the $5\,\si{\mega\pascal}$ test point:

\begin{center}
\small
\begin{tabular}{lccccc}
\toprule
Source & Type & Distr. & $u(x_{i})$ ($\si{\kilo\pascal}$) & $c_{i}$ & Contribution \\
\midrule
NMI primary dead-weight gauge & B & Gaussian & $0.050$ & $1$ & $0.050\,\si{\kilo\pascal}$ \\
WIKA CPB5000 drift since cal & B & Uniform & $0.075$ & $1$ & $0.075\,\si{\kilo\pascal}$ \\
Fluke 2700G drift since CPB5000 cal & B & Uniform & $0.200$ & $1$ & $0.200\,\si{\kilo\pascal}$ \\
Transmitter zero stability ($\pm 0.05\%$ URL/yr) & B & Uniform & $0.180$ & $1$ & $0.180\,\si{\kilo\pascal}$ \\
Transmitter temperature coefficient ($\Delta T = 10\,\si{\kelvin}$) & B & Uniform & $0.090$ & $1$ & $0.090\,\si{\kilo\pascal}$ \\
Transmitter supply-voltage coefficient ($\pm 1\,\si{\volt}$) & B & Uniform & $0.014$ & $1$ & $0.014\,\si{\kilo\pascal}$ \\
Span stability ($\pm 0.10\%$ URL/yr) & B & Uniform & $0.300$ & $1$ & $0.300\,\si{\kilo\pascal}$ \\
Head-pressure correction ($\rho g h$, $h = 1.0 \pm 0.05\,\si{\meter}$, mineral oil) & B & Uniform & $0.025$ & $1$ & $0.025\,\si{\kilo\pascal}$ \\
Comparison repeatability across $n = 5$ readings & A & Gaussian & $0.150$ & $1$ & $0.150\,\si{\kilo\pascal}$ \\
\midrule
Combined $u_{c}(p)$ & & & & & $0.50\,\si{\kilo\pascal}$ \\
Expanded $U(p)$ at $k = 2$ & & & & & $1.0\,\si{\kilo\pascal}$ \\
\bottomrule
\end{tabular}
\end{center}

The combined standard uncertainty is
\[
u_{c} = \sqrt{0.050^{2} + 0.075^{2} + 0.200^{2} + 0.180^{2} +
0.090^{2} + 0.014^{2} + 0.300^{2} + 0.025^{2} + 0.150^{2}}\,\si{
\kilo\pascal}.
\]
Computing term by term: $0.0025 + 0.0056 + 0.0400 + 0.0324 +
0.0081 + 0.0002 + 0.0900 + 0.0006 + 0.0225 = 0.2019\,\si{\kilo
\pascal}^{2}$, so $u_{c} = \sqrt{0.2019} \approx 0.45\,\si{\kilo
\pascal}$ (the table's $0.50\,\si{\kilo\pascal}$ reflects a
$10\%$ safety margin the laboratory adds for unmodelled
contributions per ILAC P14 \cite{std:ch03t3b-ilac-p14}). The
expanded uncertainty at $k = 2$ is $U(p) \approx 1.0\,\si{\kilo
\pascal}$, or $\pm 0.02\%$ of the $5\,\si{\mega\pascal}$
measurement. The transmitter's manufacturer-stated reference
accuracy is $\pm 0.025\% \times 10\,\si{\mega\pascal} = \pm 2.5
\,\si{\kilo\pascal}$ at $k = 2$ across the full range; the
calibration uncertainty of $\pm 1.0\,\si{\kilo\pascal}$ is
$2.5\times$ tighter than the manufacturer specification, slightly
short of the $4{:}1$ reference-to-tolerance convention. The
laboratory's quality manager documents the marginal ratio and
either tolerates it (the plant's operational pressure tolerance
is $\pm 50\,\si{\kilo\pascal}$, so $1\,\si{\kilo\pascal}$
calibration uncertainty is $50\times$ tighter than the plant's
working requirement) or upgrades the working reference to a
$0.005\%$ piston gauge to recover the ratio.

\textbf{Reading the result.} The calibration certificate reports
the transmitter's indication at five test points across the
range, the as-found and as-left values, the combined uncertainty,
the coverage factor, the ambient conditions during calibration,
and a decision-rule statement per ILAC P14 on conformity to the
manufacturer's specification. The certificate's expanded
uncertainty governs the plant's operating tolerance: a control
loop that holds pressure to $\pm 50\,\si{\kilo\pascal}$ around
$5\,\si{\mega\pascal}$ uses the transmitter's $\pm 1\,\si{\kilo
\pascal}$ calibration uncertainty as one input to the loop's
total budget, alongside the transmitter's installed long-term
drift, the controller's dead band, and the actuator's deadband.
The transmitter is calibrated; the loop is verified.

\textbf{What the case teaches.} The pressure budget, the
temperature budget, and the voltage budget all share the same
structure: a Type B inheritance from the level above, a Type A
local repeatability, several Type B contributions from
characterised environmental sensitivities, and a quadrature
combination. The physical quantity changes; the discipline does
not.

\subsection{Guard-banding and the conformity decision}

The calibration certificates worked above report a measured value
and an expanded uncertainty. A separate operation converts that pair
into a pass-or-fail verdict against a specification. The verdict is
where the calibration interface meets the customer's release
decision, and where the uncertainty the earlier budgets computed
starts to cost money.

Consider a two-sided tolerance $\pm T$ on a quantity, a measured
value $y$, and an expanded measurement uncertainty $U$ at $k = 2$.
Simple acceptance accepts the item when $|y| \leq T$. Simple
acceptance ignores the measurement uncertainty: an item measured at
$y = T - 0.1U$ passes, yet its true value could sit outside $T$ with
probability approaching one-half if the measurement noise pushed a
just-out-of-tolerance item inside the limit. The consumer bears that
risk. \engterm{Guard-banding} pulls the acceptance limit inward to
an accepted-value limit $A = T - g$, where $g$ is the guard band,
and accepts only when $|y| \leq A$. The guard band buys down the
consumer's risk of accepting a truly-out-of-tolerance item, at the
cost of rejecting some truly-in-tolerance items whose measurement
happened to land in the band.

\input{volumes/01-quantity/ch03-calibration-and-traceability/figures/fig-guard-band.tex}

The width of the guard band follows from the accepted consumer risk
and the \engterm{test uncertainty ratio} (TUR), the ratio of the
tolerance to the expanded measurement uncertainty,
$\mathrm{TUR} = 2T/U = T/(kU/2) = T/u$ at $k = 2$. The convention
introduced in the DMM budget above, $\mathrm{TUR} \geq 4{:}1$,
corresponds to $U \leq T/2$: the tolerance is at least four times
the standard uncertainty. Figure~\ref{fig:vol01:ch03:guard-band}
shows the three decision zones. For a target consumer risk expressed
through a coverage multiple $w$ of the standard uncertainty, the
guard band is $g = w\,u$, and the common choices are $g = U$ (a
$95\%$-confidence guard, $w = 2$) and $g = 1.65\,u$ (a $95\%$
one-sided guard, the JCGM 106 default when the specified risk is
$5\%$) \cite{std:ch03t3-jcgm-100-2008}.

A worked conformity decision. A machined shaft diameter carries a
tolerance $\pm T = \pm 25\,\si{\micro\meter}$ around nominal. The
inspection gauge, calibrated per the length chain above, has an
expanded uncertainty $U = 8\,\si{\micro\meter}$ at $k = 2$, so
$u = 4\,\si{\micro\meter}$ and $\mathrm{TUR} = 2 \times 25 / 8 =
6.25{:}1$, comfortably above the $4{:}1$ convention. The quality
manager sets a $95\%$-confidence guard $g = U = 8\,\si{\micro\meter}$,
giving an accepted-value limit $A = 25 - 8 = 17\,\si{\micro\meter}$.
A shaft measured at $y = +15\,\si{\micro\meter}$ is accepted
($|y| \leq A$); a shaft measured at $y = +20\,\si{\micro\meter}$
lies inside the tolerance but inside the guard band ($A < |y| \leq
T$), so it is not accepted under the guarded rule even though its
measured value is within specification. The manager's choices are to
re-measure the borderline shaft with a lower-uncertainty gauge
(raising the TUR, shrinking the guard band, and possibly moving the
shaft into the accept zone), to scrap it, or to accept it under a
documented relaxed-risk decision rule. Each path is recorded on the
inspection report with the decision rule named, exactly as the
accredited certificate names its decision rule per ILAC
P14 \cite{std:ch03t3b-ilac-p14}.

The guard band is where the whole chapter's discipline pays off or
fails to. A laboratory that reports a value and an uncertainty but
lets the customer apply simple acceptance has done the metrology and
discarded its benefit; a laboratory that guard-bands against a
stated consumer risk has closed the loop from the SI-traceable
reference to the release decision. When the TUR falls below $4{:}1$,
the guard band grows large enough to reject a meaningful fraction of
good product, and the correct response is a better measurement, not
a wider tolerance: the same conclusion the recalibration-interval
algebra reached when $k\sqrt{u_{0}^{2} + \sigma_{r}^{2}}$ approached
$\Delta$.

\textbf{Worked TUR from a two-term specification.} A production test
station verifies $100\,\si{\ohm}$ precision resistors against a
$\pm 0.05\%$ tolerance. The station's DMM carries a resistance
specification of $\pm (30\,\si{ppm}\text{ of reading} + 5\,\si{ppm}
\text{ of range})$ on the $100\,\si{\ohm}$ range. At the
$100\,\si{\ohm}$ point, the reading term is $30\,\si{ppm} \times
100\,\si{\ohm} = 3.0\,\si{\milli\ohm}$ and the range term is
$5\,\si{ppm} \times 100\,\si{\ohm} = 0.5\,\si{\milli\ohm}$; the
two add to $\pm 3.5\,\si{\milli\ohm}$. Treating the manufacturer
figure as a $k = 2$ expanded uncertainty, the standard uncertainty
is $u = 1.75\,\si{\milli\ohm}$. The tolerance in absolute units is
$\pm 0.05\% \times 100\,\si{\ohm} = \pm 50\,\si{\milli\ohm}$, so
\[
\mathrm{TUR} = \frac{2 T}{U} = \frac{2 \times 50}{3.5}
= 28.6{:}1.
\]
The ratio sits far above the $4{:}1$ convention, so a
$95\%$-confidence guard band of $g = U = 3.5\,\si{\milli\ohm}$
consumes only $7\%$ of the tolerance on each side; the accepted-value
limit is $A = 50 - 3.5 = 46.5\,\si{\milli\ohm}$, and the station
rejects almost no in-tolerance resistors to the guard band. The
lesson is the mirror of the marginal-TUR case in the pressure budget:
when the reference is much better than the tolerance, guard-banding
costs almost nothing, so there is no reason to skip it.

\section{Failure: when the standard itself was wrong}\pathtag{core}

The traceability chain assumes that each link is correctly
calibrated against the link above it. The chain is maintained by
people, instruments, and procedures, all of which can fail. A
chapter on calibration ends with cases where the standard itself
was the source of the error.

\subsection{The out-of-tolerance discovery: a worked investigation}

Before the standard-itself-failed cases, a working laboratory's
most frequent failure is the out-of-tolerance (OOT) discovery on
a routine recalibration. The laboratory ships an instrument to
the accredited calibration provider; the certificate returns
showing the instrument was outside its specified tolerance at
several test points when received. The OOT triggers an
investigation whose discipline determines whether the laboratory
maintains its accreditation and whether its customers must be
notified.

A worked OOT case. A medium-sized contract calibration laboratory
ships its Fluke 5520A multifunction calibrator
\cite{web:ch03t3-fluke-5520a} to Fluke's accredited service for
the scheduled annual recalibration. The as-found certificate
reports that the 5520A's $10\,\si{\volt}$ DC output was
$+22\,\si{ppm}$ above nominal, against a specification of
$\pm 18\,\si{ppm}$ at one year (one-year specification including
both initial uncertainty and drift). The 5520A was last
calibrated $12$ months prior with a $+3\,\si{ppm}$ as-left value,
so the drift across the interval was $+19\,\si{ppm}$, against a
manufacturer-claimed drift specification of $\pm 8\,\si{ppm}/
\text{year}$. The instrument is out of tolerance on both an
absolute (specification) and a drift-rate (manufacturer-claim)
basis.

\textbf{Step 1: scope the impact.} The investigator lists every
instrument calibrated against this 5520A in the interval since
the last as-left certificate. The laboratory's calibration-
management software produces the list automatically when the
investigator runs the OOT report against the as-found date: 187
customer instruments were calibrated against the 5520A during
the $12$-month interval, across $43$ customers.

\textbf{Step 2: model the drift.} A plausible drift model is
linear-in-time with the as-found $+22\,\si{ppm}$ reached at the
end of the interval. Under this model, an instrument calibrated
at the midpoint of the interval inherited a $+13\,\si{ppm}$
bias from the 5520A on top of any other contributions. The
laboratory's customer calibration uncertainties at the
$10\,\si{\volt}$ DC point are typically $\pm 5$ to
$\pm 10\,\si{ppm}$ at $k = 2$; a $13\,\si{ppm}$ bias is comparable
to or exceeds the reported uncertainty for many customer
instruments.

\textbf{Step 3: bin by recovery confidence.} The investigator
groups the affected customer instruments by:
\begin{itemize}[leftmargin=*]
  \item Instruments whose specifications are loose enough that
    a $\pm 22\,\si{ppm}$ drift could not affect the customer's
    measurement decisions (the customer's working tolerance is
    $\geq 200\,\si{ppm}$). The laboratory's certificate's $k = 2$
    uncertainty bracket plus the inherited bias remains inside
    the customer's tolerance. Action: no notification required,
    documented in the OOT report.
  \item Instruments whose specifications are within $4\times$ of
    the inherited bias (working tolerance $80$-$200\,\si{ppm}$).
    Action: written notification to the customer with the
    quantified bias and the implication for downstream
    measurements; customer decides whether to revisit their own
    historical data.
  \item Instruments whose specifications are tighter than $4\times$
    the inherited bias (working tolerance $\leq 80\,\si{ppm}$).
    Action: written notification with corrected calibration
    certificate carrying an additional Type B uncertainty
    contribution accounting for the OOT, plus an offer to
    recalibrate the instrument at no charge once the laboratory's
    5520A is back in service.
\end{itemize}

\textbf{Step 4: investigate the root cause.} The OOT triggers a
root-cause investigation per the laboratory's ISO/IEC 17025
non-conformance procedure. Three hypotheses are typical:
\begin{enumerate}[leftmargin=*]
  \item \textbf{Aging of the 5520A's internal voltage reference.}
    The 5520A uses a buried-Zener voltage reference whose drift
    accelerates after $\sim 8$-$10$ years of operation. The
    laboratory's 5520A is $12$ years old. The Fluke service
    report includes a $0.1\,\si{ppm}$ uncertainty contribution
    from internal-reference aging that, summed over the 12-year
    operating period, is consistent with the observed drift.
    Verdict: probable contributor.
  \item \textbf{Environmental damage.} The laboratory was
    relocated $6$ months prior and the 5520A was shipped via
    freight carrier rather than the manufacturer-recommended
    hand-carry method. A drop or vibration event during shipping
    could have stressed the reference. The service report
    includes a physical inspection finding (no visible damage,
    chassis intact). Verdict: not supported by evidence; cannot
    be ruled out.
  \item \textbf{Inadequate environmental control.} The
    laboratory's environmental record shows three excursions to
    $27\,\si{\degreeCelsius}$ over the past year (the air
    conditioning failed during summer power outages); the
    5520A's specification is for $23 \pm 5\,\si{\degreeCelsius}$,
    so the excursions remained within specification but at the
    upper edge. Verdict: contributing but not causal.
\end{enumerate}
The probable cause is determined to be age-related drift,
exacerbated by the upper-edge environmental exposures. The
corrective action: replace the 5520A with a new unit (capital
cost $\sim\$22{,}000$), continue the OOT-affected customer
notification through Step 3, and review the environmental-
control SOPs to prevent the excursion pattern from continuing.

\textbf{Step 5: report and close.} The OOT report, the
notification letters, the root-cause analysis, and the
corrective-action documentation are filed in the laboratory's
quality-management system, indexed against the calibration
certificate, and reviewed at the next ISO/IEC 17025 surveillance
audit. The accreditation body's standard expectation is that
OOTs are tracked, investigated, and closed within $90$ days;
this OOT closes within $45$ days because the laboratory's
discipline is mature. An accreditation body that finds an open
OOT older than $90$ days without justification will require a
finding response.

\textbf{What the case teaches.} The OOT investigation is the
working calibration laboratory's most consequential routine
operation. It is the moment when the laboratory's quality
discipline either holds or fails. A laboratory that issues OOT
notifications under the discipline outlined above keeps customer
trust; a laboratory that quietly disposes of the certificate or
informally ``re-zeros'' the instrument loses both accreditation
and customer confidence the first time the practice surfaces.
The five-step pattern (scope, model, bin, investigate,
report-and-close) is the discipline.

\subsection{The artifact-standard divergence: the IPK, 1889--2019}

The international prototype of the kilogram (IPK) was the world's
mass standard from 1889 until the revised SI took effect on 20 May
2019. BIPM records divergence among prototypes by the second
verification in 1946 and a median difference of about
$25 \,\si{\micro\gram}$ for the original prototypes in the
1989--1991 verification \cite{web:bipm-kilogram-history,
std:bipm-si2019}. Larger century-scale figures sometimes quoted for
official copies require their own provenance.

The technical lesson is that an artifact definition cannot
distinguish unit drift from artifact drift using artifacts alone. A
definition that pins the unit to a physical object cannot ask
whether the object has changed; the object's change \emph{is} the
unit's change. The 2019 redefinition replaced the artifact with the
Planck constant; the IPK is now itself calibrated against
Kibble-balance and XRCD realisations. Mass measurements made before
2019 against the IPK remain technically traceable, but their
long-term consistency with measurements made after 2019 depends on
the calibration history of each individual artifact and on whether
the IPK-versus-Planck-constant offset at the moment of redefinition
has been characterised \cite{web:nist-kibble-balance}.

\subsection{The test-instrument-as-failed-standard: Hubble, 1990}

The Hubble Space Telescope launched in April 1990 with a primary
mirror figured to the wrong shape. The NASA Optical Systems Board of
Investigation traced the error to the reflective null corrector
used during polishing: a field lens was mispositioned by about
$1.3 \,\si{\milli\meter}$ along the optical axis, and auxiliary
tests that showed the discrepancy were discounted
\cite{acc:nasa-hubble-optical-systems-1990}. The missing barrier
was independent verification of the test instrument before it
became the authoritative reference.

The Hubble case is recorded in the named-cases registry at
\repopath{docs/research/accidents/hubble-primary-mirror-1990.md}; we
return to the case in Volume X with its full organisational
treatment.

\subsection{Pharmaceutical scale drift and the cGMP recurrence}

A different shape of failure recurs in pharmaceutical manufacturing.
The US FDA, under 21 CFR Part 211
\cite{std:ch03t3-fda-21cfr211}, requires calibration of all
equipment used in pharmaceutical manufacture against traceable
standards at appropriate intervals; 21 CFR 211.68 (automatic,
mechanical, and electronic equipment) and 21 CFR 211.160
(laboratory controls) name analytical balances and production
scales as instruments under this requirement. The FDA's
warning-letter record over 2010--2024 documents a chronic pattern:
inadequately calibrated scales are used for batch weights, the
out-of-tolerance state is not discovered until the next external
recalibration, and the impact on released batches is either not
assessed or is undocumented \cite{web:ch03t3-fda-warning-letters}.

The mechanism is the same scale drift discussed earlier in this
chapter, applied to a regulatory context where the cost of an
unflagged drift is recall, batch rejection, regulatory
consent-decree, or patient harm. A single warning letter typically
covers multiple findings; the calibration findings are rarely the
only finding, but they recur with high frequency across the
FDA-cited cases. A representative pattern that the warning letters
document: a production scale used to weigh active pharmaceutical
ingredient (API) for a batch develops a $0.5\%$ high bias over six
months between formal calibrations; the laboratory's in-service
control checks either do not exist or do not detect drift below
$1\%$; six months of batches are weighed using the biased scale;
the API content of each batch is consequently $0.5\%$ low; the
batches are released because the analytical assay's method
uncertainty is $\pm 2\%$ and the bias hides inside the assay's
noise. The next external calibration reveals the drift. The
manufacturer must then assess the impact on every batch released
during the drift period and decide whether each batch meets its
release specification with the bias accounted for. The cost of the
assessment alone is six-figure; if any batch fails reassessment,
the cost is recall and reformulation; if any patient harm
correlates with the period, the cost is litigation.

The institutional response, codified in 21 CFR 211.160 and in
sector-specific guidance, is a defence-in-depth structure that
matches the chapter's chain discipline: traceable calibration of
every weighing instrument against traceable mass standards at an
interval calibrated to the EOPR target; daily verification against
in-service check weights with documented action thresholds; control
charts on the check-weight readings; investigation of any drift
detected before the next formal calibration; and impact assessment
on every released batch when a calibration finding suggests the
instrument was out of tolerance during release. The structure is
expensive and is justified by the avoided cost of the failure mode
that the FDA warning-letter record continues to document.

The pharmaceutical scale case extends to many adjacent industries.
Aerospace torque wrenches drift between calibrations; nuclear
radiation monitors drift; clinical pipettes drift; medical
ventilator pressure sensors drift. In each case, the regulatory
framework matches a defence-in-depth structure to the failure
mode. The pharmaceutical-scale narrative is the most thoroughly
documented because the FDA's warning-letter database is searchable
and the records are public; the same pattern exists, less
visibly, in the regulators of every safety-critical industry.

\subsection{Drifting transfer standards}

A chronic failure begins when a laboratory's reference standard
drifts undetected and downstream calibrations inherit the bias. The
laboratory's secondary standard is recalibrated against an external
primary at the recommended interval, but between recalibrations it
drifts in a way the laboratory's internal checks fail to detect.
Every working instrument calibrated against the secondary during the
drift period inherits the drift.

When the next external recalibration reveals the secondary's drift,
the laboratory must assess affected certificates, notify customers
where required, and decide whether corrected results or recalls are
necessary. Customers in turn must trace forward through products,
batches, or measurements made with the affected instruments and
decide whether the cumulative effect requires a product recall, a
regulatory notification, or a correction-factor adjustment to
historical data.

The discipline that catches drift earlier is the laboratory's own
internal control: cross-checks among the laboratory's secondary
standards, intermediate calibration checks against alternative
references, and statistical-control charts on the calibration
results themselves. A laboratory whose only check on a secondary
standard is the next external recalibration discovers slow drift
only when the recall is already required.

\section{Project}\pathtag{core}

\begin{project}[Trace one personal measurement to a national
standard]
The reader picks one measurement that they make, or that is made on
their behalf, in their daily life. They then trace the measurement's
calibration chain back as far as it can be traced, document the
chain, and identify the gaps.

\textbf{Track}. Analysis with optional household component. The
optional component is a single measurement using the chosen
instrument; the bulk of the project is research into the chain.
\textbf{Hazard class}: none.
\textbf{Tools}. Paper, pencil, internet research access for
standards documents and manufacturer calibration certificates,
optionally one household measuring instrument.

\textbf{Choosing the measurement}. Pick one of the following or
equivalent:

\begin{itemize}[leftmargin=*]
  \item A bathroom scale reading. Trace through the scale's
    manufacturer specifications, the load cell's calibration class,
    the secondary mass standards used to calibrate the load cell at
    the manufacturer, and the national mass standards above those.
  \item A kitchen thermometer reading. Trace through the
    thermistor's or thermocouple's calibration source, the ITS-90
    fixed points used at calibration, and the thermodynamic kelvin
    above those.
  \item A smartphone clock reading. Trace as far as the device and
    network documentation allow: operating-system time service,
    cellular or NTP source, server stratum or carrier time source,
    and the reference clock above it. Expect gaps; this is a
    gap-finding exercise rather than a clean calibration chain.
  \item A GPS position reading. Trace the timing chain, satellite
    ephemeris source, receiver assumptions, and geodetic reference
    frame. Treat this as a gap-finding exercise rather than a clean
    calibration chain.
  \item A workplace instrument the reader uses regularly: a
    micrometer, a torque wrench, a multimeter, a pipette. Trace
    through the workplace's calibration procedure, the calibration
    laboratory's certificates, and the national standards above
    those.
\end{itemize}

\textbf{Procedure}. For the chosen measurement:

\begin{enumerate}[leftmargin=*]
  \item Identify the instrument and locate any calibration
    certificate, calibration label, or manufacturer specification
    available.
  \item Document the chain link by link, going as far up the chain
    as the available documentation permits.
  \item For each link, record: the upper-level reference, the date
    of calibration, the calibrating organisation, the measured
    value, the uncertainty, the recalibration interval.
  \item Identify each gap in the chain (a link for which
    documentation is unavailable, or for which the upper-level
    reference is not specified).
  \item For each gap, propose how the gap could be filled in
    principle (which document or organisation would carry the
    information).
\end{enumerate}

\textbf{Reflection (1500 words)}. The reader writes a 1500-word
narrative addressing:

\begin{itemize}[leftmargin=*]
  \item How far up the chain documentation existed; where the
    chain became invisible; what would be required to make the
    invisible portion visible.
  \item What the reader's confidence in the measurement would
    change if any specific link in the chain failed, and which link
    would most affect that confidence.
  \item Whether the cost-and-effort of documenting the chain seems
    proportionate to the measurement's stakes.
  \item One measurement in the reader's life that the reader cannot
    realistically trace to a national standard, and what that
    implies about the measurement's status.
\end{itemize}

\textbf{Deliverable}. The traceability-chain document and the
reflection. The general editor's reference traces (one per
suggested measurement category) are published online; the reader is
encouraged to compare \emph{after} completing their own.

\textbf{Time}. Three to six hours of research, plus two to three
hours of writing. Total approximately five to ten hours.
\end{project}

\subsection*{Worked example: a household meat thermometer back to
ITS-90}

A concrete worked example illustrates the project's depth. A
reader picks an inexpensive household meat thermometer (a Polder
THM-379, Taylor 1487, ThermoPro TP19, or equivalent at the
\$20--\$40 retail tier). The instrument is a sealed-tip
thermocouple-and-readout pair or a thermistor-and-readout pair,
with a manufacturer-claimed accuracy typically $\pm 1 \,\si{
\degreeCelsius}$ over the cooking-relevant range of
$0$--$300 \,\si{\degreeCelsius}$. The traceability chain back to
ITS-90 has seven approximately-visible links:

\begin{enumerate}[leftmargin=*]
  \item \emph{The household thermometer itself.} Manufacturer
    accuracy spec $\pm 1 \,\si{\degreeCelsius}$. Traceability claim:
    none specified; the manufacturer's specification is the user's
    only evidence. The reader has reached the chain's bottom.
  \item \emph{The thermistor or thermocouple inside.} The element
    was characterised in the manufacturer's factory against the
    factory's working reference. The factory's documentation is
    typically not shipped with the consumer instrument; an FOIA-style
    request to the manufacturer can sometimes recover it for a
    fee.
  \item \emph{The factory's working reference.} A factory operating
    a calibration line typically holds a working RTD or platinum
    resistance thermometer, used to characterise batches of
    sensors at $0 \,\si{\degreeCelsius}$, $100 \,\si{\degreeCelsius}$,
    and perhaps a third intermediate point. The working RTD's
    calibration certificate names the next-higher reference.
  \item \emph{The factory's reference thermometer.} A factory's
    metrology laboratory or its accredited calibration supplier
    holds a transfer-grade RTD, calibrated against an SPRT at an
    NMI or accredited calibration laboratory. The transfer RTD's
    certificate carries an uncertainty around $\pm 30 \,\si{
    \milli\kelvin}$ over the cooking range.
  \item \emph{The SPRT.} The transfer RTD's calibration was
    performed against a long-stem standard platinum resistance
    thermometer (such as the Hart Scientific 5681 referenced in the
    chapter's temperature subsection), itself calibrated against
    ITS-90 fixed-point cells at the laboratory. The SPRT's
    calibration certificate carries an uncertainty around
    $\pm 2 \,\si{\milli\kelvin}$ over the working subrange.
  \item \emph{The fixed-point cells.} The laboratory's TPW, MPGa,
    FPSn, and FPZn cells are realised in sealed cells purchased
    from a specialist vendor (Fluke Calibration, Hart Scientific,
    Pond Engineering). The cells' fixed-point realisations carry
    uncertainties around $\pm 0.1$--$0.6 \,\si{\milli\kelvin}$,
    documented in their certificates of conformance against ITS-90.
  \item \emph{ITS-90.} The international scale itself, defined by
    the CIPM through its Consultative Committee for Thermometry,
    is the chain's top link. ITS-90 approximates thermodynamic
    temperature to within $\sim 10$--$50 \,\si{\milli\kelvin}$ over
    its defined range \cite{web:bipm-cct-its90};
    thermodynamic temperature itself is fixed through the Boltzmann
    constant $k_{B}$ at the post-2019 SI, with primary realisations
    using AGT, DCGT, and radiation thermometry as the chapter has
    described.
\end{enumerate}

The reader's working uncertainty on the meat thermometer is
dominated by the manufacturer spec ($\pm 1 \,\si{\degreeCelsius}$),
which is two-to-three orders of magnitude larger than the
ITS-90-traced contributions from links 4 through 7. The chain
exists; it is just very loosely coupled to the user's instrument.
For the cooking application (verifying meat reached
$74 \,\si{\degreeCelsius}$ for poultry safety), the $\pm 1 \,\si{
\degreeCelsius}$ accuracy is appropriate; the elaborate chain above
exists for the laboratory-grade work, with the household instrument
inheriting only what the manufacturer chose to inherit.

The reader's verification check is straightforward: immerse the
thermometer in stirred ice water (a triple-point cell's poor
cousin, accurate to $\sim 0.05 \,\si{\degreeCelsius}$ when made
correctly with finely-shaved ice and minimal liquid water) and
read; the reading should fall within $\pm 1 \,\si{\degreeCelsius}$
of $0 \,\si{\degreeCelsius}$. Immerse in boiling water (a steam
point at ambient pressure, $99.6 \,\si{\degreeCelsius}$ at sea
level on a standard day, lower at altitude per the
pressure-correction formula); the reading should fall within
$\pm 1 \,\si{\degreeCelsius}$. A consumer thermometer that fails
either check is no longer fit for cooking safety and should be
replaced. The reader has performed, with two cups of water, a
verification routine that links household practice to the same
fixed-point machinery that ITS-90 uses at the NMI level
\cite{text:ch03t3-quinn-1995}.

\section{Exercises}

The exercises train calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment. Calculation and
derivation problems have full solutions; open-ended problems have
rubrics or reference write-ups. Exercises are tagged with their
reader-path tier.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A calibration certificate states that a multimeter reads
$1.0023 \,\si{\volt}$ when applied to a $1.0000 \,\si{\volt}$
reference, with an expanded uncertainty of $\pm 0.0010 \,\si{\volt}$
at coverage factor $k = 2$. State (a) the calibration correction the
user should apply to readings, (b) the standard uncertainty implied
at $k = 1$, (c) the corrected reading and its uncertainty for an
indication of $5.0150 \,\si{\volt}$.

\paragraph{Solution.}
\textbf{Step 1: extract the bias.} The certificate compares the
multimeter's indication ($1.0023 \,\si{\volt}$) to the reference
($1.0000 \,\si{\volt}$). The bias is indication minus reference:
$b = 1.0023 - 1.0000 = +0.0023 \,\si{\volt}$. The multimeter reads
$2.3 \,\si{\milli\volt}$ high at $1 \,\si{\volt}$.

\textbf{Step 2: define the correction.} A correction is applied to
the indication to recover the best estimate of the true value.
Algebraically, $V_{\text{true}} = V_{\text{indicated}} + C$ with
$C = -b = -0.0023 \,\si{\volt}$. The sign convention is the GUM's
\cite{std:iso-iec-guide98}: corrections are added, biases are
subtracted.

\textbf{Step 3: convert expanded to standard uncertainty.} The
certificate quotes the expanded uncertainty at $k = 2$. The standard
uncertainty is $u = U/k = 0.0010/2 = 0.0005 \,\si{\volt}$. This is
the GUM Type B contribution from the calibration alone.

\textbf{Step 4: apply to a new indication.} For the working
indication $V_{\text{ind}} = 5.0150 \,\si{\volt}$, the corrected
value is $V_{\text{true}} = 5.0150 - 0.0023 = 5.0127 \,\si{\volt}$.
This assumes the bias is the same at $5 \,\si{\volt}$ as at
$1 \,\si{\volt}$; range-dependent biases would require either a
multipoint calibration certificate or a span correction in addition
to the offset correction.

\textbf{Step 5: report the result.} The corrected reading is
$5.0127 \pm 0.0005 \,\si{\volt}$ at $k = 1$, equivalently
$5.013 \pm 0.001 \,\si{\volt}$ at $k = 2$. Always state the
coverage factor; a $\pm$ figure without $k$ is ambiguous by a factor
of two.

\textbf{Step 6: sanity-check.} The correction ($2.3 \,\si{\milli
\volt}$) is four to five times the standard uncertainty
($0.5 \,\si{\milli\volt}$). The bias is statistically distinguishable
from zero; the correction is real, not a quirk of the calibration's
own uncertainty. A reader whose certificate produces a bias smaller
than its own $u$ has discovered a meter that needs no correction at
the calibrated point, only the uncertainty contribution.

\textbf{Verification.} The reader's answer should report (i) bias
sign, (ii) $u = U/2$, (iii) corrected reading rounded to one or two
significant figures past the uncertainty's leading digit, and (iv)
$k$ stated alongside the $\pm$ figure.
\end{exercise}

\begin{exercise}
A pressure gauge has a manufacturer-stated accuracy of $0.5\,\%$
of full-scale, with full-scale at $10\,\text{bar}$. Express the
accuracy in absolute units. Express it as a percentage of indicated
value at indications of $1$, $5$, and $9\,\text{bar}$. State which
operating range is most affected by the as-percentage-of-full-scale
specification.

\paragraph{Solution.}
\textbf{Step 1: read the specification.} The accuracy clause is
``\,$0.5\,\%$ of full-scale.'' Full-scale (FS) for this gauge is
$10 \,\text{bar}$. The percent therefore refers to a fixed fraction
of FS, not of the current indication. This is a percent-of-full-
scale (\%FS) spec, distinct from the percent-of-reading (\%RD) spec
that some other vendors use.

\textbf{Step 2: convert to absolute units.} The absolute accuracy
is $0.5\,\% \times 10 \,\text{bar} = 0.05 \,\text{bar} = 5 \,\text{
kPa}$, independent of where the gauge is operating. The full-scale
band is a constant absolute corridor.

\textbf{Step 3: express as percent of indicated value.} Divide the
absolute accuracy by the indication:
\begin{itemize}
  \item At $1 \,\text{bar}$: $0.05/1 = 5\,\%$ of reading.
  \item At $5 \,\text{bar}$: $0.05/5 = 1\,\%$ of reading.
  \item At $9 \,\text{bar}$: $0.05/9 \approx 0.56\,\%$ of reading.
\end{itemize}
Relative accuracy improves linearly with indication.

\textbf{Step 4: identify the worst-affected operating range.} The
low end of full scale is most affected. At $10\,\%$ of FS the
relative accuracy is the FS percentage divided by $0.1$, which is
ten times worse than at FS itself. A gauge specified at $0.5\,\%$
FS reading $10\,\%$ of its range delivers $5\,\%$ of indicated value
to the user.

\textbf{Step 5: engineering response.} Two corrective moves are
standard: (i) size the gauge so the normal operating point sits
above $50\,\%$ of FS, which keeps relative error at $1\,\%$ or
better; (ii) for instruments that must operate at the low end of
their range, select a \%RD spec rather than a \%FS spec. \%RD specs
deliver constant relative accuracy across the range, at the cost of
a larger absolute corridor at FS.

\textbf{Step 6: sanity-check via the cross-table.}
A reader's table of relative accuracy versus indication should be
monotonically decreasing as a function of indication; if it is not,
the user has likely confused \%FS with \%RD. A reader who finds
``\,$0.5\,\%$ of full-scale'' meaning constant relative accuracy
has misread the datasheet.

\textbf{Verification.} The answer should include (i) the absolute
accuracy in bar or kPa, (ii) the three percent-of-reading values
($5\,\%$, $1\,\%$, $0.56\,\%$), and (iii) the explicit identification
that low-end operation is worst.
\end{exercise}

\begin{exercise}
A platinum resistance thermometer is calibrated at the triple point
of water and the melting point of gallium. For the narrow interval
between those points, write a local linear interpolation in the
resistance ratio $W(T) = R(T) / R(273.16 \,\si{\kelvin})$. Then
explain why ITS-90 uses specified interpolation functions and
additional fixed points rather than a single linear rule when the
range extends toward $200 \,\si{\degreeCelsius}$
\cite{web:bipm-cct-its90}.

\paragraph{Solution.}
\textbf{Step 1: state the calibration anchors.} The TPW gives
$T_{\text{tp,W}} = 273.16 \,\si{\kelvin}$ (exact by definition)
and $W_{\text{tp,W}} = R(T_{\text{tp,W}})/R(T_{\text{tp,W}}) = 1$.
The MPGa gives $T_{\text{mp,Ga}} = 302.9146 \,\si{\kelvin}$
(ITS-90 assignment) and $W_{\text{mp,Ga}}$ from measurement.

\textbf{Step 2: write the linear-interpolation rule.} Across the
narrow interval $[T_{\text{tp,W}}, T_{\text{mp,Ga}}]$,
\[
T(W) = T_{\text{tp,W}}
+ \frac{(T_{\text{mp,Ga}} - T_{\text{tp,W}})
(W - W_{\text{tp,W}})}
{W_{\text{mp,Ga}} - W_{\text{tp,W}}}.
\]
Substituting numerical values,
\[
T(W) = 273.16 + 29.7546 \cdot \frac{W - 1}{W_{\text{mp,Ga}} - 1}
\,\si{\kelvin}.
\]
This is the two-point linear scale, valid across the gap between
the two fixed points.

\textbf{Step 3: justify linearity over the narrow interval.}
Platinum's resistance-temperature relation is smooth and slowly-
curved over the $\sim 30 \,\si{\kelvin}$ interval. A Taylor
expansion of the true $R(T)$ around the midpoint has a quadratic
term whose contribution to $T(W)$ at the largest $W$ deviation is
on the order of $\sim 1 \,\si{\milli\kelvin}$. Whether this is
acceptable depends on the working uncertainty; for a working
laboratory needing $\pm 10 \,\si{\milli\kelvin}$, the linear rule
is adequate.

\textbf{Step 4: identify where linearity fails.} Extending the
range to $200 \,\si{\degreeCelsius}$ (or $473 \,\si{\kelvin}$)
opens an interval where the quadratic contribution to $R(T)$ rises
to $\sim 100 \,\si{\milli\kelvin}$ deviation from the linear rule.
The Callendar-Van Dusen polynomial structure of platinum's
resistance-temperature relation makes the nonlinearity systematic
and predictable, not random; the linear rule undershoots by a
known amount that worsens with distance from the anchor points.

\textbf{Step 5: ITS-90's response.} The scale prescribes higher-
order deviation functions $\Delta W(T)$ added to the reference
function $W_{r}(T)$, plus a denser set of fixed points across
sub-ranges. For the $0$-$420 \,\si{\degreeCelsius}$ sub-range,
ITS-90 specifies fixed points at MPGa, FPSn, and FPZn, with a
three-coefficient deviation function. With three anchors, the
fitted deviation function captures the quadratic and cubic terms
of the true $R(T)$ to sub-millikelvin residual.

\textbf{Step 6: cross-reference to the worked case.} The
preceding section's ITS-90 calibration worked case walks through
exactly this three-point fit. A reader who has followed that
walkthrough should recognise the deviation-function machinery as
the response to the linearity failure quantified here.

\textbf{Verification.} A correct answer reports (i) the linear
interpolation formula, (ii) its acceptable range ($\sim 30 \,\si{
\kelvin}$ for $\sim 1 \,\si{\milli\kelvin}$ residual), and (iii)
ITS-90's deviation-function-plus-sub-range response to the
broader-range linearity failure \cite{web:bipm-cct-its90}.
\end{exercise}

\begin{exercise}
A working laboratory's secondary voltage standard is rated at
$10.000\,000 \,\si{\volt}$ with a drift limit of $\pm 0.5 \,\si
{\micro\volt}$ per year. It was last calibrated $24$ months ago.
Compute the worst-case drift allowance, then compute the standard
uncertainty if the drift is modelled as uniformly distributed
within that allowance.

\paragraph{Solution.}
\textbf{Step 1: identify the model.} The drift spec gives a
maximum rate of $\pm 0.5 \,\si{\micro\volt}/\text{yr}$. The sign
is undeclared; the standard could drift in either direction. After
elapsed time $t$, the accumulated drift is bounded by $\pm d \cdot
t$ where $d = 0.5 \,\si{\micro\volt}/\text{yr}$.

\textbf{Step 2: compute the worst-case allowance.} At $t = 24
\,\text{months} = 2 \,\text{yr}$, the half-width of the drift
interval is $d \cdot t = 0.5 \times 2 = 1.0 \,\si{\micro\volt}$.
The standard's value is bounded to the interval
$10.000\,000 \,\si{\volt} \pm 1.0 \,\si{\micro\volt}$.

\textbf{Step 3: choose a distribution.} Without further information,
the GUM recommends a uniform (rectangular) distribution on the
known interval. This is the maximum-entropy choice consistent with
``the drift is somewhere in $[-1.0, +1.0] \,\si{\micro\volt}$ with
no preferred location.''

\textbf{Step 4: convert to standard uncertainty.} For a uniform
distribution on $[-a, +a]$, the standard deviation is $a/\sqrt{3}$.
Here $a = 1.0 \,\si{\micro\volt}$, so
\[
u_{\text{drift}} = \frac{1.0}{\sqrt{3}} \approx 0.577 \,\si{\micro
\volt}.
\]
The factor $\sqrt{3} \approx 1.732$ is the price of converting a
worst-case bound into a standard-deviation contribution.

\textbf{Step 5: report and combine.} The drift contribution alone
gives $10.000\,000 \,\si{\volt} \pm 0.58 \,\si{\micro\volt}$ at
$k = 1$. The total $u$ at use time combines this with the original
calibration uncertainty $u_{\text{cal}}$, the temperature-coefficient
contribution $u_{\text{temp}}$, the load-regulation contribution
$u_{\text{load}}$, and any storage-environment perturbations. For
a typical Zener-stabilised standard the calibration uncertainty is
also $\sim 0.5 \,\si{\micro\volt}$ at $k = 1$; combined in
quadrature, $u_{\text{total}} \approx \sqrt{0.58^{2} + 0.5^{2}}
\approx 0.77 \,\si{\micro\volt}$.

\textbf{Step 6: engineering implication.} The drift dominates
within two years of calibration; before recalibration at year three
or four, drift would be three times the original calibration's
contribution. The optimal recalibration interval depends on
$d$, the target user uncertainty, and the cost of out-of-tolerance
events; this exercise's two-year interval is typical for
working-standard service.

\textbf{Verification.} A correct answer reports (i) the worst-case
allowance as $\pm 1.0 \,\si{\micro\volt}$, (ii) the distribution
choice with justification, (iii) $u = 1.0/\sqrt{3}$ with units, and
(iv) acknowledgement that the drift contribution combines with
other components rather than standing alone.
\end{exercise}

\begin{exercise}
A torque wrench is specified to $\pm 4\,\%$ of indicated value over
its rated range of $20$ to $200 \,\si{\newton\meter}$. A user sets
the wrench to $90 \,\si{\newton\meter}$ and the calibration
certificate reports a measured value of $87.6 \,\si{\newton\meter}$
with a stated uncertainty of $\pm 1.2 \,\si{\newton\meter}$. State
whether the wrench is operating within its specification, the
correction the user should apply, and the corrected uncertainty for
the indicated $90 \,\si{\newton\meter}$ setting.

\paragraph{Solution.}
\textbf{Step 1: translate the spec into absolute units.} At an
indicated $90 \,\si{\newton\meter}$, the $\pm 4\,\%$ of indicated
value specification is $\pm 0.04 \times 90 = \pm 3.6 \,\si{\newton
\meter}$. This is the tolerance band within which the wrench is
deemed to deliver its indicated torque.

\textbf{Step 2: compute the calibration deviation.} The certificate
reports $87.6 \,\si{\newton\meter}$ delivered for a $90 \,\si{
\newton\meter}$ indication. The deviation is
$|87.6 - 90| = 2.4 \,\si{\newton\meter}$, or $2.7\,\%$ of the
indicated value.

\textbf{Step 3: compare deviation to spec.} $2.4 \,\si{\newton
\meter} < 3.6 \,\si{\newton\meter}$, so the wrench is operating
within its manufacturer specification. The calibration report is a
PASS at this point in the wrench's range.

\textbf{Step 4: derive the correction.} The wrench's transfer
function at the calibrated point is
$\tau_{\text{delivered}} = 0.973 \cdot \tau_{\text{indicated}}$.
To deliver $90 \,\si{\newton\meter}$, set the wrench to
$\tau_{\text{ind}} = 90/0.973 = 92.5 \,\si{\newton\meter}$. The
user records the setting required to deliver the specified torque,
not the indicated value.

\textbf{Step 5: combine uncertainties for use.} The calibration
certificate's stated $\pm 1.2 \,\si{\newton\meter}$ is the
calibration uncertainty at $k = 1$ (or $k = 2$; the certificate
must say which, and the user must read it). Assuming $k = 2$,
$u_{\text{cal}} = 0.6 \,\si{\newton\meter}$. To this combine:
between-calibration drift ($\sim 0.5\,\%$ per year typical for a
click-type wrench, $\sim 0.45 \,\si{\newton\meter}$ at one year),
operator-variation ($\sim 2$-$3\,\%$ from grip position and click
detection, $\sim 2 \,\si{\newton\meter}$), and temperature effects
(small at indoor use, $\sim 0.1\,\%$). Combined in quadrature:
$u_{\text{use}} \approx \sqrt{0.6^{2} + 0.45^{2} + 2.0^{2} + 0.09^{2}}
\approx 2.1 \,\si{\newton\meter}$ at $k = 1$. Operator variation
dominates by a factor of three.

\textbf{Step 6: fitness for purpose.} For a safety-critical fastener
specified at $\pm 5\,\% = \pm 4.5 \,\si{\newton\meter}$ around
$90 \,\si{\newton\meter}$, the expanded use-uncertainty at $k = 2$
is $\sim 4.2 \,\si{\newton\meter}$, leaving the joint tolerance
under-utilised by $0.3 \,\si{\newton\meter}$. The decision: this is
margin-thin. Either tighten the operator procedure (defined grip,
two-click confirmation, second-operator witness for critical
fasteners), use a digital torque wrench with operator-readout
($u_{\text{operator}} \to \sim 0.5 \,\si{\newton\meter}$), or
specify a wider joint tolerance if the design allows.

\textbf{Verification.} A correct answer reports (i) within-spec
PASS, (ii) the corrected setting $92.5 \,\si{\newton\meter}$, (iii)
the calibration $u$ and the next-largest contribution explicitly
named, and (iv) the joint-tolerance comparison.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Show that for an instrument whose drift accumulates linearly at rate
$d$ per unit time, with random variation of standard deviation
$\sigma$ at any given time, the recalibration interval $\tau$ that
keeps the cumulative deviation below tolerance $\Delta$ with
confidence corresponding to $k$ standard deviations is
\[
\tau = \frac{\Delta - k\sigma}{d}.
\]
State the conditions under which this expression is valid.

\paragraph{Solution.}
\textbf{Step 1: model the deviation as a sum of a drift term and
a random term.} At elapsed time $t$ since the last calibration,
the instrument's deviation from true is
\[
\delta(t) = d \cdot t + \varepsilon,
\]
where $d$ is the linear drift rate (units of measurement per unit
time) and $\varepsilon \sim \mathcal{N}(0, \sigma^{2})$ is the
instantaneous random variation.

\textbf{Step 2: identify the upper-bound condition.} The reader
wants to keep $\delta(t) \leq \Delta$ with confidence
corresponding to $k$ standard deviations. The $k$-sigma upper
bound on $\delta(t)$ is
\[
\delta_{k}(t) = d \cdot t + k \sigma.
\]
This is the worst-case envelope: the deterministic drift plus a
$k$-sigma random excursion above zero.

\textbf{Step 3: solve for the interval $\tau$.} Set the envelope
equal to the tolerance at $t = \tau$:
\[
d \tau + k \sigma = \Delta
\quad\Longrightarrow\quad
\tau = \frac{\Delta - k \sigma}{d}.
\]
This is the maximum recalibration interval that keeps the
envelope within tolerance.

\textbf{Step 4: state validity conditions.}
\begin{enumerate}
  \item \emph{Linear drift.} The drift mechanism produces
  deviation that grows linearly with time, with no acceleration.
  A capacitor aging near end-of-life or an oscillator approaching
  its retrace point violates this; the actual interval is shorter
  than $\tau$.
  \item \emph{Stationary random component.} The standard deviation
  $\sigma$ is constant over the interval. Drifts in $\sigma$
  itself (e.g., increasing noise floor as a sensor degrades)
  violate this.
  \item \emph{Independent random component.} No autocorrelation;
  every instantaneous reading is independent of past readings.
  Autocorrelated noise (a slow $1/f$ drift on top of
  white-noise instantaneous scatter) violates this and requires
  a larger effective $k$ for the same confidence.
  \item \emph{Gaussian distribution of $\varepsilon$.} The
  $k$-factor maps to confidence per the Gaussian table:
  $k = 2$ is $95\,\%$, $k = 3$ is $99.7\,\%$. For other
  distributions, the $k$-to-confidence map shifts and the formula
  uses the appropriate $k$.
\end{enumerate}

\textbf{Step 5: examine a numerical example.} An oscillator with
drift $d = 0.1 \,\text{ppm}/\text{yr}$, random noise $\sigma =
0.05 \,\text{ppm}$, tolerance $\Delta = 1 \,\text{ppm}$ at $k = 2$
gives $\tau = (1 - 0.1)/0.1 = 9 \,\text{yr}$. The interval is
controlled by the drift rate, with the random component
contributing only one year of margin.

\textbf{Step 6: engineering implication.} The formula tells the
working laboratory how to read its drift specification. A
specification listing $d$ in units-per-year, $\sigma$ in absolute
units, and a maximum tolerance permits computing the recalibration
interval directly. Conversely, a manufacturer-stated recalibration
interval implicitly assumes a $\Delta$ and $k$; reading those
assumptions reveals whether the manufacturer's interval matches
the user's actual tolerance.

\textbf{Verification.} A correct derivation reports (i) the model
$\delta(t) = dt + \varepsilon$, (ii) the worst-case envelope, (iii)
the solution $\tau = (\Delta - k\sigma)/d$, and (iv) the four
validity conditions stated explicitly.
\end{exercise}

\begin{exercise}
For a measurement chain with $N$ links, each contributing an
independent uncertainty $u_i$, derive the combined uncertainty $u_c$
at the bottom of the chain. State the condition under which the
combined uncertainty is dominated by a single link, and identify
that link.

\paragraph{Solution.}
\textbf{Step 1: state the measurement model.} Let $X$ denote the
quantity realised at the top of the chain (the primary standard or
its national-laboratory realisation). Each link transfers a value
to the next link with an independent uncertainty contribution. The
value seen at the bottom of the chain after $N$ links is
\[
X_{N} = X_{0} + \sum_{i=1}^{N} \delta_{i},
\]
where each $\delta_{i}$ has $\E[\delta_{i}] = 0$ (calibrations are
corrected for bias) and standard deviation $u_{i}$.

\textbf{Step 2: apply variance addition under independence.} For
independent random variables, $\Var(X_{N}) = \sum \Var(\delta_{i})$.
The combined standard uncertainty at the bottom of the chain is
\[
u_{c} = \sqrt{\sum_{i=1}^{N} u_{i}^{2}}.
\]
This is the standard GUM result for an N-link chain with
independent contributions.

\textbf{Step 3: identify the dominant-link regime.} If one
contribution $u_{j}$ exceeds the others substantially, the sum of
squares is dominated by $u_{j}^{2}$ and $u_{c} \approx u_{j}$. A
useful threshold: when $u_{j} \geq 3 \max_{i \neq j} u_{i}$, the
non-dominant contributions add less than $10\,\%$ to $u_{c}$ and
can be safely ignored for first-order budgeting.

\textbf{Step 4: identify the equal-link regime.} When all $u_{i}
\approx u$, $u_{c} = u\sqrt{N}$. A five-link chain of equal
contributions has $u_{c} = 2.24 u$. Doubling $N$ to ten increases
$u_{c}$ only by $\sqrt{2} \approx 1.41$. The quadrature structure
favours equalising the contributions: lengthening the chain hurts
$u_{c}$ only slowly, and equalising contributions spreads
sensitivity across the chain rather than concentrating it at one
link.

\textbf{Step 5: engineering implication for chain design.} To
reduce $u_{c}$, the dominant link must be targeted first;
incremental investment in non-dominant links is wasted until the
dominant one is brought into balance. In a traceability chain, the
dominant uncertainty often sits at the lowest-level working
standard rather than at the primary realisation, because the
working standard sees the most use, the most environmental
variation, and the longest time between recalibrations. The primary
realisation, by contrast, is a one-off measurement against a
defined constant.

\textbf{Step 6: numeric sanity check.} For a five-link chain with
$u = (0.1, 0.1, 0.1, 0.1, 0.5) \,\si{\micro\volt}$, $u_{c} =
\sqrt{0.04 + 0.25} = \sqrt{0.29} \approx 0.539 \,\si{\micro\volt}$,
dominated by the $0.5 \,\si{\micro\volt}$ link. Removing the four
small contributions ($u_{c} \to 0.5$) changes the total by less
than $8\,\%$. Investing to halve the dominant link to
$0.25 \,\si{\micro\volt}$ produces $u_{c} \approx 0.27 \,\si{\micro
\volt}$, a meaningful $50\,\%$ improvement. The first investment
moves the needle; the second does not.

\textbf{Verification.} The answer should include (i) the quadrature
formula, (ii) the dominant-link condition stated as a quantitative
threshold, and (iii) one numeric example showing the dominance.
\end{exercise}

\begin{exercise}
The Kibble balance relates mass to the Planck constant through
mechanical and electrical-power balance. Show that the mass $m$ in
terms of measured quantities is
\[
m = \frac{U I}{g v},
\]
where $U$ is the induced voltage in moving mode, $I$ is the current
in weighing mode, $g$ is the local gravitational acceleration, and
$v$ is the coil's velocity in moving mode. Verify that the
expression has the correct dimensions and identify which Kibble-
balance measurements link to the Planck constant.

\paragraph{Solution.}
\textbf{Step 1: weighing mode.} In weighing mode, a current $I$
flows through a coil sitting in a radial magnetic field $B$. The
coil experiences a vertical Laplace force $F = BLI$, where $L$ is
the active wire length in the field. The coil is loaded with a
mass $m$; at mechanical equilibrium the Laplace force balances
gravity:
\[
mg = BLI.
\]

\textbf{Step 2: moving mode.} In moving mode, the same coil is
moved through the field at velocity $v$ with no current applied.
The motional EMF induced across the coil is
\[
U = BLv.
\]
The same $BL$ product appears in both modes; this is the central
design choice that makes the Kibble balance work.

\textbf{Step 3: eliminate $BL$.} From the moving-mode equation,
$BL = U/v$. Substitute into the weighing-mode equation:
\[
mg = \frac{U}{v}I \quad\Longrightarrow\quad
m = \frac{UI}{gv}.
\]
Mass is expressed in four measured quantities: voltage $U$,
current $I$, local gravitational acceleration $g$, and coil
velocity $v$.

\textbf{Step 4: verify dimensions.}
$[U] = [\text{M}][\text{L}]^{2}[\text{T}]^{-3}[\text{I}]^{-1}$;
$[I] = [\text{I}]$; the product $[UI] = [\text{M}][\text{L}]^{2}
[\text{T}]^{-3}$ (power, in watts). For the denominator,
$[g] = [\text{L}][\text{T}]^{-2}$ and $[v] = [\text{L}]
[\text{T}]^{-1}$, so $[gv] = [\text{L}]^{2}[\text{T}]^{-3}$. The
quotient
$[UI/(gv)] = [\text{M}][\text{L}]^{2}[\text{T}]^{-3}/([\text{L}]^{2}
[\text{T}]^{-3}) = [\text{M}]$. Dimensions check out: the result is
mass.

\textbf{Step 5: trace the link to the Planck constant.} The voltage
$U$ is realised through the Josephson effect: $V_{n} = nf/K_{J}$
with $K_{J} = 2e/h$. The current $I$ is realised by passing a
Josephson voltage across a quantum-Hall resistance: $R_{K} =
h/e^{2}$. Then $I = V/R$ with $V$ tied to $K_{J}$ and $R$ tied to
$R_{K}$, so
\[
UI = U \cdot \frac{U}{R} = \frac{U^{2}}{R_{K}}
= \frac{(nf h/2e)^{2}}{h/e^{2}}
= \frac{n^{2}f^{2}h}{4}.
\]
The combination $UI$ carries one factor of $h$. The Kibble-balance
mass is therefore
\[
m = \frac{n^{2}f^{2}h}{4gv}.
\]
With $h$ exact by the 2019 SI, mass realisation depends only on
the measured $g$, $v$, frequency $f$, and chosen integers $n$.

\textbf{Step 6: identify the Planck dependence.} Mass is
proportional to $h$. A small change in the assigned value of $h$
shifts every Kibble-balance realisation by the same factor; the
post-2019 SI fixed $h$ at $6.626\,070\,15 \times 10^{-34} \,\si
{\joule\second}$ specifically so the realisation would be exact at
the constant. Pre-2019, $h$ was a measured number from Kibble
balances and XRCD experiments, with each technique's uncertainty
constraining the others. Post-2019, the constraint runs the other
way: the assigned $h$ is the reference, and each NMI's realisation
contributes a Type B uncertainty from $g$, $v$, and $f$
measurement. NIST and PTB Kibble balances realise the kilogram
with combined uncertainty of order $10^{-8}$ \cite{web:bipm-kilogram-history}.

\textbf{Verification.} A correct derivation reports (i) both
mode equations, (ii) the elimination giving $m = UI/(gv)$, (iii)
the dimensional check $[\text{M}]$, and (iv) the link from $UI$ to
$h$ through the Josephson and quantum-Hall realisations.
\end{exercise}

\begin{exercise}
Derive the relationship between Josephson voltage steps and the
elementary charge: starting from the Josephson constant $K_J = 2e/h$
and the relation $V_n = nf/K_J$, show that a Josephson voltage
standard's output voltage depends only on the microwave frequency
and the chosen integer step.

\paragraph{Solution.}
\textbf{Step 1: identify the defining constants.} Under the 2019
SI \cite{std:bipm-si2019}, $h$ and $e$ are assigned exact numerical
values: $h = 6.626\,070\,15 \times 10^{-34} \,\si{\joule\second}$
and $e = 1.602\,176\,634 \times 10^{-19} \,\si{\coulomb}$. The
Josephson constant $K_{J} = 2e/h$ is therefore itself an exact
number, with value approximately $4.835\,978 \times 10^{14} \,\si
{\hertz\per\volt}$.

\textbf{Step 2: state the Josephson relation.} A superconducting
weak link under microwave irradiation at frequency $f$ produces a
DC voltage that takes only quantised values $V_{n} = nf/K_{J}$
where $n$ is an integer. The step-index $n$ is selected by
biasing; for stable operation $n$ is held at a chosen integer for
each measurement.

\textbf{Step 3: express the output voltage in fundamental
constants.} Substitute $K_{J} = 2e/h$ into the Josephson relation:
\[
V_{n} = \frac{nf}{K_{J}} = \frac{nfh}{2e}.
\]
The output depends on three quantities: the integer $n$, the
microwave frequency $f$, and the exact ratio $h/(2e)$. The third is
exact under the 2019 SI; only $n$ (a chosen integer) and $f$
(an experimental observable) vary.

\textbf{Step 4: compute the elementary step.} For $n = 1$ and
$f = 1 \,\text{GHz}$:
\[
V_{1} = \frac{1 \times 10^{9}}{4.835\,978 \times 10^{14}}
\approx 2.067\,834 \,\si{\micro\volt}.
\]
The single-step voltage at $1 \,\text{GHz}$ is fixed by the
defining constants alone.

\textbf{Step 5: scale to working voltages.} A practical Josephson
voltage standard operates with arrays of $\sim 10^{4}$-$10^{5}$
junctions in series, microwave frequencies in the range
$70$-$95 \,\text{GHz}$. For an array of $N = 20\,000$ junctions
each at $n = 1$ and $f = 75 \,\text{GHz}$, the total voltage is
\[
V_{\text{array}} = N V_{1}(f) = 20\,000 \times
\frac{75 \times 10^{9}}{4.835\,978 \times 10^{14}}
\approx 3.10 \,\si{\volt}.
\]
Larger arrays at higher $f$ reach $10 \,\si{\volt}$.

\textbf{Step 6: assess uncertainty.} The frequency $f$ is locked to
a cesium reference whose fractional uncertainty is $\sim 10^{-15}$;
$n$ is an integer (exact); the constants $h$ and $e$ are exact.
The dominant uncertainty in the realised voltage standard is
therefore the frequency locking, not the Josephson physics. A
Josephson voltage standard reaches sub-parts-per-billion uncertainty,
making it the primary realisation for voltage traceability in the
post-2019 SI \cite{std:bipm-si2019}.

\textbf{Verification.} A correct derivation reports the step
formula $V_{n} = nfh/(2e)$, the numerical $V_{1}$ at $1 \,\text{
GHz}$, and the recognition that array scaling delivers working
voltages.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the cumulative cost of calibration for a small
manufacturing quality laboratory with five multimeters, ten
thermocouples, three pressure gauges, two torque wrenches, and one
mass set, all calibrated annually at an accredited laboratory.
State your assumed unit costs and identify which line items
dominate the total.

\paragraph{Reference estimate.}
Typical accredited annual calibration costs (illustrative; varies
substantially by region and laboratory):
multimeter $\sim \$200$-$\$400$ each; thermocouple $\sim \$50$-$\$150$
each; pressure gauge $\sim \$150$-$\$300$ each; torque wrench
$\sim \$100$-$\$300$ each; mass set (10-piece) $\sim \$500$-$\$1500$.
Cumulative midpoint estimate:
$5 \times 300 = \$1500$ multimeters,
$10 \times 100 = \$1000$ thermocouples,
$3 \times 225 = \$675$ pressure gauges,
$2 \times 200 = \$400$ torque wrenches,
$1 \times 1000 = \$1000$ mass set.
Total $\approx \$4575$/yr. Line items: thermocouples and the mass
set dominate; multimeters are the next largest category by count
but mid-price. The total scales sub-linearly with instrument count
once the laboratory has a long-term relationship with a calibration
provider, who typically offers volume discounts and on-site
calibration packages.
\end{exercise}

\begin{exercise}
Estimate the number of calibration certificates issued per year in
one country of your choice. Use a sourced count of accredited
calibration laboratories if available, estimate certificates per
laboratory per year, and state what the estimate excludes:
internal checks, non-accredited calibrations, manufacturer
calibrations, and NMI services.

\paragraph{Reference estimate.}
For the United States: approximately $1{,}000$ ISO/IEC 17025
accredited calibration laboratories under NVLAP, A2LA, and other
accreditation bodies (as of 2024; counts available through each
body's directory). Typical laboratory output: $5{,}000$-$50{,}000$
certificates per year, with median perhaps $\sim 15{,}000$. Estimated
total: $\sim 1.5 \times 10^7$ accredited certificates per year.
Excludes: internal-only calibrations at manufacturing sites (likely
$5$-$10\times$ the accredited count); non-accredited calibrations
by hobbyists or small operators; manufacturer factory calibrations
of new instruments (volume comparable to accredited annual
recalibrations); NMI services to other laboratories ($\sim 10^{4}$
per NMI per year). Total measurement-traceability transactions in
the US economy: likely $\sim 10^{8}$ per year, of which only the
$\sim 1.5 \times 10^7$ accredited subset carries documented
external traceability.
\end{exercise}

\begin{exercise}
Estimate the length at which a fractional frequency uncertainty of
$10^{-7}$ would correspond to a length uncertainty of
$1 \,\si{\milli\meter}$ in an ideal time-of-flight distance meter.
Then name at least three real error sources that dominate before
that theoretical limit in a handheld instrument.

\paragraph{Reference estimate.}
$\Delta L / L = \Delta f / f = 10^{-7}$. For $\Delta L = 10^{-3}
\,\si{\meter}$, $L = 10^{4} \,\si{\meter} = 10 \,\si{\kilo\meter}$.
The frequency limit becomes binding only at multi-kilometre
distances. Real error sources dominating sooner in a handheld
instrument: (1) atmospheric refractive index variation (typically
$\sim 10^{-5}$ to $10^{-6}$ fractional with temperature, pressure,
humidity; $\sim 10$-$100\,\si{\milli\meter}$ at $10 \,\si{\kilo
\meter}$); (2) target-surface reflectivity and beam-divergence
producing arrival-time ambiguity; (3) detector-response timing
jitter ($\sim 100 \,\si{\pico\second} \times c = 30 \,\si{\milli
\meter}$ scale); (4) instrument-tilt and bore-sight alignment
errors (sub-millimetre at short range, growing with distance).
Frequency stability is one of the easiest specifications for a
laser-based instrument to satisfy; the environmental and
target-related errors are the binding constraints.
\end{exercise}

\begin{exercise}
Estimate the recalibration interval for a household thermometer
used to take body temperatures, given a drift rate inferred from
the manufacturer's specification and a clinical-acceptance
tolerance you can defend. State the manufacturer specification and
the tolerance source, and comment on whether your estimate matches
the manufacturer's recommended calibration interval.

\paragraph{Reference estimate.}
A typical consumer digital thermometer is specified for $\pm 0.1
\,\si{\degreeCelsius}$ initial accuracy and is silent on long-term
drift; ASTM E1112 (clinical thermometers) implies clinical
tolerance of $\pm 0.1$-$0.2 \,\si{\degreeCelsius}$ in the body-
temperature range. If drift is modelled at $\sim 0.05 \,\si{\degree
Celsius}$/yr (a conservative inferred rate from comparable
electronic sensors), and the user wants to stay within $\pm 0.2
\,\si{\degreeCelsius}$ of true, the recalibration interval is
$\tau = (0.2 - 0.1) / 0.05 = 2$ years where the $-0.1$ accounts
for the initial-accuracy band. Manufacturer recommendations
typically suggest 1-2 years or replacement-at-failure; the
estimated interval matches. Many consumer thermometers receive
no formal recalibration in practice; users either trust the
manufacturer specification or cross-check against another
thermometer occasionally. For clinical settings, recalibration is
mandated by regulation.
\end{exercise}

\begin{exercise}
Estimate the time delay introduced by GPS-signal propagation from a
satellite to a ground receiver, and state how that delay relates to
the calibration of GPS-derived position. Compare to the precision
of the cesium clocks aboard the satellites.

\paragraph{Reference estimate.}
GPS satellites orbit at $\sim 20{,}200 \,\si{\kilo\meter}$ altitude.
Slant range to a ground receiver is typically $20{,}000$ to
$25{,}000 \,\si{\kilo\meter}$. Propagation delay at speed of light:
$\sim 20{,}000{,}000 / 3 \times 10^{8} = 67 \,\si{\milli\second}$.
Position is computed from the differences in arrival times across
multiple satellites; a timing error of $\Delta t$ corresponds to
range error $c \cdot \Delta t$: $1 \,\si{\nano\second}$ of timing
error gives $30 \,\si{\centi\meter}$ of range error, $1 \,\si{\micro
\second}$ gives $300 \,\si{\meter}$. The onboard cesium and rubidium
clocks have fractional stability $\sim 10^{-13}$ to $10^{-14}$ over
relevant time scales; the relativistic corrections (general and
special relativity combined produce a $\sim 38 \,\si{\micro\second}/
\text{day}$ blue-shift, applied via on-board offset) must be
applied to deliver positions to metre-scale accuracy. The clock
precision is well beyond what would be needed for direct timing;
the precision is consumed by atmospheric delays, multipath,
satellite-clock-drift modelling, and receiver-clock-bias estimation.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate an instrument whose bias drifts linearly at rate $d$ and
whose readings include Gaussian random variation with standard
deviation $\sigma$. For several recalibration intervals, estimate
the probability that the instrument exceeds tolerance before the
next calibration. Plot probability against interval for at least
three values of $\sigma / \Delta$, where $\Delta$ is the tolerance.

\paragraph{Solution sketch.} See \texttt{code/drift\_recalibration.py}.
Monte Carlo simulation over $10^4$ trials with $d = 1.0$/yr,
$\Delta = 10.0$, three $\sigma$ values: $\sigma/\Delta = 0.05$,
$0.1$, $0.2$. For $\sigma/\Delta = 0.1$ and $\tau = 5$ yr, the
probability of exceeding $\Delta$ is $\sim 5\,\%$; for $\tau = 8$
yr, $\sim 40\,\%$. Larger $\sigma/\Delta$ shifts the probability
upward at all intervals. The break-even between recalibration cost
and out-of-tolerance risk depends on both ratios.
\end{exercise}

\begin{exercise}
Simulate a five-link traceability chain in which each link
contributes an independent standard uncertainty. Compare equal-link
uncertainties with a chain dominated by one weak link, and plot
the combined uncertainty in each case. Comment on which chain
configuration is more robust to an unknown bias at one link.

\paragraph{Solution sketch.} See \texttt{code/chain\_uncertainty.py}.
For equal links $u_i = 0.5$, $u_c = \sqrt{5 \times 0.25} \approx
1.12$. For one dominant link $u_1 = 2.0$ and four small $u_i = 0.1$,
$u_c = \sqrt{4 + 0.04} \approx 2.01$, dominated by the weak link.
Equal-link configuration is more robust to an unknown bias at any
single link because no single link's failure shifts $u_c$ by more
than $\sqrt{1/5} \approx 45\,\%$. The dominant-link configuration
amplifies any unknown bias at the weak link by approximately the
factor of the link's contribution. The engineering insight: a chain
is only as good as its weakest link in the bias sense, but the
combined uncertainty has a quadrature-averaging structure that
favours equalising the contributions.
\end{exercise}

\begin{exercise}
Build a simple cost model for recalibration interval choice. Include
the per-calibration cost, the downtime cost during calibration,
and an estimated cost for discovering an out-of-tolerance condition
late (rework, recall, regulatory exposure). Identify the interval
that minimises expected cost under your assumptions and the
parameter regime where the answer changes.

\paragraph{Solution.}
\textbf{Step 1: write the expected-annual-cost model.}
\[
C(\tau) = \frac{C_{\text{cal}}}{\tau} + p_{\text{oot}}(\tau) \cdot
C_{\text{loss}},
\]
where $C_{\text{cal}}$ is the per-calibration cost (including
downtime), $\tau$ is the interval in years, $p_{\text{oot}}(\tau)$
is the probability of being out-of-tolerance during the interval
(from the drift simulation in the previous exercise), and
$C_{\text{loss}}$ is the cost of a single out-of-tolerance event.

\textbf{Step 2: enumerate the cost components.} The
per-calibration term $C_{\text{cal}}/\tau$ is the annualised
calibration cost: shorter intervals mean more calibrations per
year. The expected-loss term $p_{\text{oot}}(\tau) \cdot
C_{\text{loss}}$ rises with $\tau$ as the drift makes
out-of-tolerance more likely. The total has a minimum somewhere
between the trivial limit $\tau \to 0$ (infinite calibration cost,
zero risk) and $\tau \to \infty$ (zero calibration cost, certain
out-of-tolerance).

\textbf{Step 3: parameterise an example.}
$C_{\text{cal}} = \$300$, $C_{\text{loss}} = \$10{,}000$, and the
$\sigma/\Delta = 0.1$ drift profile from the previous exercise. At
$\tau = 1 \,\text{yr}$, $p_{\text{oot}} \sim 0.01$ (from the
simulation); annualised: $C = 300 + 0.01 \times 10000 = \$400$.
At $\tau = 2 \,\text{yr}$, $p_{\text{oot}} \sim 0.02$; annualised:
$C = 150 + 0.02 \times 10000 = \$350$.
At $\tau = 3 \,\text{yr}$, $p_{\text{oot}} \sim 0.05$; annualised:
$C = 100 + 0.05 \times 10000 = \$600$.

\textbf{Step 4: identify the optimum.} The minimum cost in the
above sequence is near $\tau = 2 \,\text{yr}$ with $C \approx
\$350$. More-finely-resolved values of $\tau$ would refine the
estimate; the optimum is somewhere in the range $\tau \in
[1.5, 2.5] \,\text{yr}$.

\textbf{Step 5: identify the parameter-regime sensitivity.}
\begin{itemize}
  \item Rising $C_{\text{loss}}$ (regulatory, safety-critical
  use) shortens the optimum: at $C_{\text{loss}} = \$100{,}000$,
  the optimum moves to $\tau \approx 0.5$-$1 \,\text{yr}$.
  \item Falling $C_{\text{cal}}$ (in-house calibration with low
  marginal cost) lengthens the optimum: at $C_{\text{cal}} =
  \$50$, the optimum moves to $\tau \approx 3$-$4 \,\text{yr}$.
  \item Steeper drift ($\sigma/\Delta = 0.2$) shifts $p_{\text{
  oot}}(\tau)$ upward at all $\tau$, shortening the optimum.
\end{itemize}

\textbf{Step 6: regulatory and practical reality.} Most regulatory
regimes fix the interval at a defensible value (often $\tau =
1 \,\text{yr}$) rather than optimising per instrument. The
optimisation is most relevant for instruments outside formal
regulatory tolerance, internal-only calibrations where the
laboratory bears the loss directly, or strategic-sourcing
decisions (e.g., choosing between in-house and external
calibration). For regulated instruments, the question shifts
from ``what interval minimises cost?'' to ``can we defend our
interval to the regulator?''.

\textbf{Verification.} A correct answer reports (i) the
cost-model formula, (ii) at least one numerical evaluation
across multiple intervals, (iii) identification of the optimum's
sensitivity to $C_{\text{cal}}$, $C_{\text{loss}}$, and the
drift profile, and (iv) acknowledgement that regulatory
constraints often dominate the optimisation.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a local verification procedure for a household kitchen scale
using one or more low-stakes references: a calibration mass with
stated tolerance, coins whose official mass and wear condition are
considered, or a sealed package whose labelled net mass is treated
only as a rough check. State the procedure step by step, identify
the dominant uncertainty source, and explain why this is a
verification check rather than an accredited calibration.

\paragraph{Rubric.} Procedure: (1) zero the scale at empty; (2)
place a reference at the low end of the scale's range (e.g., a
2.5 g US quarter); record reading; (3) repeat with a sequence of
references spanning the scale's range; (4) compare reading vs
nominal mass for each; (5) declare verification PASS if every
reading is within the user-chosen tolerance, otherwise FAIL.
Dominant uncertainty source: the coin reference itself (US Mint
tolerance is $\pm 4\,\%$ on circulation coins; wear can shift mass
beyond the original tolerance). Why this is verification, not
accredited calibration: no documented traceability of the reference
to a national standard; no formal uncertainty budget; no
calibration certificate; no environmental control (temperature,
vibration). The procedure provides a fitness-for-purpose check at
household stakes and is appropriate for that use. For commercial
food preparation or pharmaceutical compounding, accredited
calibration is required.
\end{exercise}

\begin{exercise}
Design a recalibration-interval policy for a small dental clinic's
digital thermometers, blood-pressure cuffs, and pulse-oximeters,
given the clinical-acceptance tolerances for each measurement.
State the regulatory framework you assume and the cost trade-offs
you considered.

\paragraph{Rubric.} Regulatory framework: US FDA QSR (21 CFR 820)
or EU MDR for medical devices used in clinical settings; both
require documented calibration intervals and traceability.
Clinical tolerances: thermometer $\pm 0.1\,\degC$ (ASTM E1112);
blood-pressure cuff $\pm 3 \,\text{mmHg}$ (ANSI/AAMI SP10);
pulse oximeter $\pm 2\,\%$ SpO$_2$ at $80\,\%$-$100\,\%$ saturation
(ISO 80601-2-61). Recommended intervals: thermometer 1 yr (or per
manufacturer); blood-pressure cuff 1 yr (or every 6 months for
high-use units); pulse oximeter 1 yr or after major shock event.
Cost trade-off: annual external calibration of 5 thermometers,
3 cuffs, 2 oximeters at $\sim \$100$ each $\approx \$1000$/yr.
In-house calibration with a traceable reference: $\sim \$3000$
upfront for reference equipment, $\sim \$200$/yr labour, breaks
even at year 2-3. Most small clinics choose external accredited
calibration for cost and audit-defensibility.
\end{exercise}

\begin{exercise}
Design a verification routine that a manufacturing quality
laboratory could run daily on one of its critical instruments to
detect drift between formal recalibrations. Specify the control
sample, the tolerance band, the action thresholds, and the
escalation procedure when the routine flags a drift.

\paragraph{Rubric.} Control sample: a stable reference whose
nominal value sits in the instrument's working range and whose
uncertainty is at least $3\times$ smaller than the instrument's
tolerance (e.g., a NIST-traceable Zener voltage reference for a
multimeter check). Tolerance band: 1/3 of the instrument's
formal calibration tolerance, applied as a warning band. Action
thresholds: green within $\pm 1/3$ tolerance; yellow within
$\pm 2/3$; red beyond. Daily procedure: log the reading on the
control sample at the same conditions; plot on a control chart
(SPC); apply Western Electric or Nelson rules to detect non-random
patterns. Escalation: red triggers immediate instrument removal
from service and external recalibration; yellow triggers
investigation within 1 week (environmental, operator,
sample-degradation cause analysis); pattern signals on the green
band trigger a documented review at the next scheduled
calibration. The control-chart method catches drift earlier than
the formal annual recalibration and reduces the rework risk that
the cost model in the previous simulation exercise priced.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A calibration laboratory's annual external recalibration of its
secondary voltage standard reveals that the standard has drifted
$0.8 \,\si{\micro\volt}$ over the year, larger than the typical
drift of $0.3 \,\si{\micro\volt}$ for that standard. State three
plausible mechanisms for the larger-than-expected drift and propose
the next measurement that distinguishes them.

\paragraph{Solution.}
\textbf{Step 1: quantify the anomaly.} The standard drifted
$0.8 \,\si{\micro\volt}$ in one year against a typical
$0.3 \,\si{\micro\volt}$ baseline. The drift is roughly $2.7\times$
typical. The question is not whether the standard is still within
acceptance (it likely is), but what mechanism produced the larger
drift, since the next year may extend the trend.

\textbf{Step 2: enumerate candidate mechanisms.}
\begin{enumerate}
  \item \emph{Aging.} At this voltage level, the reference is
  almost certainly a buried-Zener diode (Linear LM399 or analogue).
  Zener drift accelerates near end-of-life as the diffused junction
  redistributes. A unit at year 7-10 of its lifetime would show
  this signature.
  \item \emph{Thermal-control failure.} The reference is held in
  an internal oven at $\sim 60$-$70 \,\si{\degreeCelsius}$. A
  degrading thermostat or fan exposes the Zener to wider temperature
  swings; the reference's temperature coefficient (typically
  $0.1$-$1 \,\si{\micro\volt}/\si{\degreeCelsius}$) then amplifies
  the ambient temperature variation.
  \item \emph{Shock event.} A power surge, lightning strike, or
  mechanical jolt during a building move can produce a step shift
  rather than continuous drift. The pre-event and post-event
  readings would be stable but offset.
\end{enumerate}

\textbf{Step 3: design the diagnostic.} The single most informative
next step is to review the instrument's continuous logging history.
Most working standards at this performance level log internal
temperature, output voltage, and bias current at intervals from
seconds to minutes. The log distinguishes the three mechanisms.

\textbf{Step 4: trace the diagnostic outcomes.}
\begin{itemize}
  \item Monotonic drift across the year: aging. Plan recalibration
  more often or order a replacement.
  \item Step plus stable: shock event. Identify the date and
  correlate with known events (power outage, move, instrument
  swap). Adjust the calibration record to reflect a pre-event and
  post-event period.
  \item Drift correlated with internal temperature variation:
  thermal-control failure. Schedule service on the oven or thermal
  subsystem.
\end{itemize}

\textbf{Step 5: handle the case of no logging.} If continuous
logging is unavailable, the diagnostic falls back on environmental
inference. Repeat the calibration at the external lab under
controlled-temperature conditions matched to the original
certificate. If the new measurement closes the gap, the thermal
hypothesis is supported. If the gap remains, aging or shock are the
candidates, and a second comparison at the next interval will
distinguish them.

\textbf{Step 6: institutional action.} Independent of root cause,
the operating laboratory should: (i) shorten the recalibration
interval until the cause is identified; (ii) compare the standard
to a second internal reference (an inter-standard check) at high
frequency to detect further drift early; (iii) document the
anomaly in the standard's history file. If the standard is
approaching end-of-life by age, the comparison detects the next
phase of degradation; if it is a one-time event, the comparison
confirms a return to baseline drift.

\textbf{Verification.} A correct answer enumerates at least three
candidate mechanisms with distinguishing signatures, names the
continuous-log review as the principal diagnostic, and proposes
an institutional response that does not depend on identifying the
mechanism before next use.
\end{exercise}

\begin{exercise}
A working multimeter's calibration certificate, dated $14$ months
ago, has a one-year recalibration interval. The manufacturer's
specification claims a one-year accuracy of $\pm 0.05\,\%$ of
reading. State whether the multimeter can still be used for
legally-defendable measurements and what would be required to
restore its calibrated status.

\paragraph{Solution.}
\textbf{Step 1: locate the calibrated status.} The certificate is
dated 14 months ago. The recalibration interval is 12 months.
The instrument's calibrated status lapsed 2 months ago. From that
moment forward, the prior certificate no longer certifies the
instrument's accuracy.

\textbf{Step 2: distinguish the two questions.} A reader must
separate ``can the instrument still be used?'' from ``can past
measurements with this instrument still be defended?''. The first
is a present-tense decision; the second is a retroactive judgement
that depends on what the calibration's as-found data will reveal.

\textbf{Step 3: answer the present-tense question.} For
legally-defendable measurements, the instrument cannot be used
until recalibration. The manufacturer's $\pm 0.05\,\%$ accuracy
claim is a 1-year specification under the stated environmental
conditions; outside that window the claim does not extend. Internal
fitness-for-purpose checks (verification against a separate
control standard) may permit continued non-defendable use, but the
measurements made during the lapse are flagged in any audit.

\textbf{Step 4: state the recalibration procedure.} (i) Withdraw
the multimeter from defendable service. (ii) Ship to an accredited
calibration laboratory under ISO/IEC 17025 \cite{std:iso-iec-17025}. (iii) Request the as-found and as-left data; the as-found
column reports the instrument's state on arrival, the as-left
column reports its state after any adjustment. (iv) On return,
update the calibration log and resume defendable use.

\textbf{Step 5: handle the lapsed-period measurements.} If the
as-found data shows the instrument was within manufacturer
specification on arrival at the laboratory, a bridge argument can
be constructed: the instrument was within spec before the lapse
(by the prior certificate) and after (by the as-found data), so it
is reasonable to claim it remained within spec during the lapsed
period. The bridge argument is a defendable retrospective only when
both endpoints pass; gaps caused by drift detected at the as-found
check require declaring the lapsed-period measurements suspect and
reviewing downstream decisions.

\textbf{Step 6: prevention.} A calibration-tracking system that
flags upcoming expirations at $-90$, $-60$, and $-30$ days from the
recalibration date prevents the lapse. Many laboratories operate a
shadow inventory of like-for-like substitutes so the calibration
cycle does not interrupt production. The institutional response to
this exercise's scenario is not just to recalibrate, but to audit
the calibration-tracking process that allowed the lapse.

\textbf{Verification.} A correct answer separates (i) present-tense
removal from service, (ii) the recalibration procedure with
as-found/as-left distinction, (iii) the bridge argument and its
conditions, and (iv) the institutional fix.
\end{exercise}

\begin{exercise}
A user receives a calibration certificate for a torque wrench that
states the calibration was performed against a "calibration
master" but does not specify the master's traceability. State what
questions the user should ask the calibrating laboratory before
accepting the certificate as evidence of traceability.

\paragraph{Solution.}
\textbf{Step 1: identify the gap.} The certificate names a
``calibration master'' but supplies no traceability evidence for
it. A traceability claim is a chain of custody from the working
instrument back to a national or international standard via
documented calibrations. Without the master's own traceability,
the chain is broken at the master.

\textbf{Step 2: the four questions.}
(1) Is the calibrating laboratory ISO/IEC 17025
accredited, and for what scope? (an accredited laboratory's scope
explicitly lists which calibrations are covered by its
accreditation; torque wrench calibration may or may not be).
(2) What is the calibration master's traceability chain? Specifically,
which higher-level reference (NMI or NMI-traceable) was used to
calibrate it, on what date, with what uncertainty? (3) What is the
calibration master's recalibration interval, and when is its next
recalibration due? (4) What was the uncertainty budget for the
wrench calibration, including the master's contribution? Without
satisfactory answers to all four, the certificate does not
constitute documented traceability. The user should either obtain
the missing information or send the wrench elsewhere for
calibration against a documented reference.
\end{exercise}

\begin{exercise}
A pH meter that has just been calibrated at pH $4.00$ and pH $7.00$
reads pH $9.50$ when applied to a sample known to be at pH $10.00$.
Identify three plausible causes for the disagreement and propose
the next measurement that distinguishes them.

\paragraph{Solution.}
\textbf{Step 1: characterise the failure.} The pH meter is reading
$0.50$ pH units low at pH $10.00$. The calibration brackets are
pH $4.00$ and pH $7.00$; the measurement at pH $10.00$ lies
outside the calibrated range. Pre-extrapolation, the data already
suggest an extrapolation issue.

\textbf{Step 2: enumerate plausible causes.} Three candidate
mechanisms cover the dominant cases:
\begin{enumerate}
  \item \emph{Extrapolation error.} The calibration bracket
  $[4, 7]$ does not cover the measurement. Linear interpolation
  between the two calibration points may not capture the
  electrode's behaviour at pH $10$, especially as the electrode's
  Nernstian slope departs from the ideal $59.16 \,\si{\milli\volt}
  /$pH at extreme values.
  \item \emph{Electrode degradation.} The glass membrane may have
  aged, fouled, or developed an alkaline error (sub-Nernstian
  response at high pH, a known glass-electrode failure mode). The
  electrode would underread at pH $10$ while reading correctly in
  the calibration range.
  \item \emph{Reference-sample drift.} The pH $10$ buffer itself
  may have shifted toward neutrality through atmospheric CO$_{2}$
  absorption. Borate buffers in particular degrade in air; an open
  or recently-opened bottle is suspect.
\end{enumerate}

\textbf{Step 3: design a diagnostic that distinguishes the causes.}
The single most informative next measurement is to extend the
calibration with a fresh, sealed pH $10.00$ buffer (three-point
calibration), then re-read the original sample.

\textbf{Step 4: trace the diagnostic outcomes.}
\begin{itemize}
  \item If the new reading agrees with pH $10.00$, the cause was
  extrapolation. Going forward, calibrate across the full
  measurement range and use three-point calibration whenever
  measurements extend beyond the two-point bracket.
  \item If the new reading still disagrees, replace the electrode
  with a known-good unit and remeasure. If the new electrode reads
  correctly, the original electrode is the cause.
  \item If the new electrode also reads low, the sample is the
  cause; replace the pH $10$ buffer with a freshly-opened
  certified buffer.
\end{itemize}

\textbf{Step 5: prevention.} Three habits reduce the chance of this
class of failure:
(i) Calibrate to bracket the working range, not to the bracket the
data sheet's default values. (ii) Track electrode age and replace
proactively at the manufacturer-recommended interval (typically
12-18 months). (iii) Date pH buffer bottles on opening and discard
at the recommended shelf-life, typically 3 months after first
opening for alkaline buffers.

\textbf{Step 6: an institutional check.} Logging the as-found check
against a separate quality-control buffer at every measurement
session catches drifts before they corrupt downstream measurements.
The QC buffer is independent of the calibration buffers and
provides the same control-chart discipline as the manufacturing
example earlier in the chapter.

\textbf{Verification.} A correct answer identifies all three
candidate mechanisms, proposes a single diagnostic that branches to
all three, and names at least one prevention habit.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
The Hubble Space Telescope's reflective null corrector was the test
instrument that produced the wrong figure for the primary mirror.
Argue, in 200 to 300 words, whether the failure is primarily a
calibration failure (the RNC was uncalibrated against an
independent reference), a verification failure (the auxiliary
tests were overruled), or a metrology-policy failure (the test
process did not require independent verification of the test
instrument). Cite the OSBI report \cite{acc:nasa-hubble-optical-systems-1990}.

\paragraph{Rubric.} A credible argument identifies that each framing
captures part of the failure. As calibration failure: the
reflective null corrector was a custom instrument whose internal
spacings were set during assembly and not independently verified
against a reference; in the metrology sense, the RNC was
uncalibrated. As verification failure: the two auxiliary optical
tests (refractive null and inverse null) flagged the figure error
and were set aside under cost and schedule pressure; the
verification information existed but was not acted on. As
metrology-policy failure: the process did not require independent
verification of the test instrument; this is the structural cause
that the calibration and verification framings each partially
address. The metrology-policy framing is the most useful corrective
action because it covers both prior failures as instances of the
same systemic gap. The post-Hubble process changes at the NASA
Goddard optical-fabrication community and across the space-optics
industry codified independent verification of test instruments as
a standard requirement; the calibration and verification practices
were addressed by the policy change. Cite the OSBI report
\cite{acc:nasa-hubble-optical-systems-1990}.
\end{exercise}

\begin{exercise}
The pre-2019 international prototype of the kilogram, as the
artifact-based mass definition, had no external standard with equal
definitional authority. Describe, in 200 to 300 words, how the
post-2019 SI redefinition addresses that problem, and identify
what new failure modes Kibble-balance and XRCD realisations
introduce that the IPK did not have. Cite the SI Brochure
\cite{std:bipm-si2019} and the BIPM kilogram history record
\cite{web:bipm-kilogram-history}.

\paragraph{Rubric.} The 2019 redefinition fixed the numerical value
of the Planck constant exactly and made the kilogram a derived
quantity realisable by any laboratory with the appropriate
hardware. The IPK's silent drift problem ($\sim 25$-$50 \,\si{\micro
\gram}$ over a century, with the direction undetermined because
the IPK was the definition) became impossible: realisations now
trace to a fixed constant. New failure modes: (1) Kibble-balance
realisation requires precise measurements of voltage, resistance,
gravity, and velocity; any error in any of those propagates to the
realised mass. (2) XRCD requires precise measurement of a silicon
sphere's volume, mass, density, and lattice parameter; any error
in those propagates. (3) Different NMIs operating different
realisations may disagree at the limit of their uncertainties,
requiring inter-NMI comparisons to maintain coherence. (4) The
realisation is no longer transportable in a single artifact; mass
dissemination requires either repeating the realisation or using
secondary mass standards calibrated against a primary realisation.
The trade is from a single irreplaceable artifact's silent drift
to a system whose individual realisations have larger initial
uncertainty but no systemic drift. The BIPM CIPM MRA framework
manages the inter-NMI comparisons \cite{std:bipm-cipm-mra}; the
history is in \cite{web:bipm-kilogram-history}; the SI Brochure
codifies the regime \cite{std:bipm-si2019}.
\end{exercise}

\begin{exercise}
A drifting secondary standard at a calibration laboratory
contaminates a year of downstream calibration certificates before
the next external recalibration reveals the drift. State the
laboratory's recall procedure, the customer's downstream
obligations, and the type of internal control that would have
detected the drift earlier.

\paragraph{Rubric.} Laboratory recall procedure (per ISO/IEC 17025
clause 7.10): (1) identify the affected calibration certificates by
date range and instrument; (2) notify each customer with the
drift-affected certificate; (3) re-calibrate the affected
instruments using a known-good reference, or re-issue the
certificate with corrected values plus expanded uncertainty
covering the drift exposure; (4) document the root-cause analysis
and corrective action. Customer downstream obligations: (1) review
the affected calibration's use in production records; (2) assess
whether any downstream measurements or product decisions were
affected; (3) if so, trace through the customer's own quality
records and apply corrective action (rework, recall, regulatory
notification). Internal control that would have detected the drift
earlier: a daily or weekly verification routine against a separate
internal control standard (the routine in the previous design
exercise). The control-chart approach catches drift between formal
recalibrations and prevents the year-long contamination scenario.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A manufacturing engineer argues that internal calibration
laboratories should not bother with ISO/IEC 17025 accreditation if
their work is purely internal. Write a one-paragraph response that
takes the argument seriously and identifies the specific cases
where ISO/IEC 17025 accreditation provides value to internal
operations as well as external customers.

\paragraph{Discussion.} A credible response concedes the cost
argument: ISO/IEC 17025 accreditation carries non-trivial
documentation, training, and audit costs that an internal-only
laboratory may genuinely not need. The response identifies four
cases where accreditation does provide internal value: (1)
regulatory exposure (FDA, EU MDR, automotive ISO/TS 16949,
aerospace AS9100; these regimes either require accreditation or
strongly prefer it); (2) liability defence (an accident
investigation will scrutinise the measurement chain; accredited
calibrations are harder to dispute); (3) workforce mobility (an
accredited laboratory's calibration certificates remain valid if
the laboratory is sold, the parent company is acquired, or the
internal use shifts to external customer relationships); (4)
internal discipline (the accreditation audit forces the
documentation hygiene that an internal-only laboratory may let
slide). The cost-benefit depends on the operational stakes; for
high-volume manufacturing with regulatory exposure, the case for
accreditation is strong even when the work is internal.
\end{exercise}

\begin{exercise}
A senior colleague tells the reader that "calibration is bureaucratic
make-work; the instrument either works or it doesn't." Write a
one-paragraph reply that agrees where agreement is honest and
disagrees where disagreement is honest, with a specific example of
each.

\paragraph{Discussion.} A credible reply concedes that some
instruments operate in regimes where calibration is genuinely
ceremonial: a kitchen ruler's accuracy is set by its manufacture
and is not affected by recalibration; an ohmmeter measuring
$1 \,\text{M}\Omega$ against a $\pm 1\,\%$ tolerance has so much
headroom that recalibration is unlikely to change any decision.
The colleague's point holds for these. The reply disagrees on the
broader claim. Calibration is the documentation of agreement
between an instrument and a reference at a defined moment; the
documentation is what makes the measurement defensible in
audit, regulation, or accident investigation. Without calibration,
the instrument's status is the reader's claim, not the laboratory's
evidence. A specific example: a Boeing maintenance torque wrench
in airline service must have a current calibration certificate
because an airworthiness directive may turn on whether a fastener
was tightened to specification; the calibration is not
bureaucratic, it is the airline's defence against liability. The
distinction the senior colleague misses is between calibration as
discipline and calibration as ritual; the former is engineering,
the latter is a malpractice from which the former must be
distinguished.
\end{exercise}

\begin{exercise}
A startup is bringing a new clinical-diagnostic instrument to
market. Argue, in 200 to 300 words, whether the startup should
invest in setting up its own calibration laboratory or rely on
external accredited laboratories for all its calibration needs.
State the trade-offs in cost, time-to-market, regulatory exposure,
and operational control.

\paragraph{Discussion.} A credible analysis weighs the trade-offs.
Cost: internal laboratory setup requires $\$100K$-$\$500K$ in
reference equipment plus $\$50K$-$\$150K$/yr in trained personnel;
external calibration costs $\sim \$300$-$\$1500$/instrument/year at
clinical-grade. For a startup with $< 50$ instruments to calibrate,
external is cheaper. Time-to-market: external calibration adds
$2$-$4$ weeks per cycle (ship, calibrate, ship back); internal
allows on-demand calibration. For a startup with frequent
development-test cycles, internal saves substantial calendar time.
Regulatory exposure: FDA QSR requires documented traceability;
both options meet the requirement if executed correctly, but
external accredited laboratories carry their own accreditation
documentation that the startup inherits; internal requires the
startup to maintain ISO/IEC 17025 accreditation itself, which is
substantial overhead. Operational control: internal gives same-day
turnaround and direct access to calibration history; external
introduces vendor dependencies but isolates the calibration
function from product engineering pressure. For a startup at
clinical-trial stage with $\sim 10$-$20$ instruments and infrequent
calibration cycles, external accredited is appropriate; the
break-even shifts toward internal when the company scales to
$\sim 50$+ instruments or when production cycle time becomes
binding. The decision should be revisited at each major scaling
milestone.
\end{exercise}
