\chapter{The three laws of thermodynamics}
\label{vol04:ch01}

\epigraph{The production of motive power is then due, in steam-engines,
not to an actual consumption of caloric, but to its transportation from
a warm body to a cold body, that is, to its re-establishment of
equilibrium.}{Sadi Carnot, \emph{Reflections on the Motive Power of
Fire} (1824), in the Mendoza translation \cite[p.~7]{hist:carnot1824}.}

\chapmeta{Half-life: foundational. Archetypes: balance, scaling. Exercise target: 38. Project track: household instrumentation. Hazard class: low (kitchen-scale temperatures and pressures). Reader path default: standard, with sections explicitly tagged. Named cases: none required in this chapter; perpetual-motion proposals are described by mechanism rather than by named example.}

\noindent
Thermodynamics is the engineering of energy accounting under
constraint. The first law fixes what is conserved; the second law fixes
which way the conserved quantity can move; the third law fixes the
zero. The chapter develops the three laws as bookkeeping rules,
identifies the state functions on which the bookkeeping depends, and
closes with the diagnostic for proposals that violate the rules. The
remaining chapters of the volume apply these rules to gases and
liquids (Ch 2), to engines and refrigerators (Ch 3), to entropy in its
microscopic form (Ch 4), and to heat and mass transfer (Ch 5-7).

\section{Energy as accounting}\pathtag{core}

A system is anything we draw a boundary around. The boundary may be
real (the metal wall of a calorimeter, the skin of a pipe) or
fictitious (a control volume in a flow, a thermodynamic envelope
around a building). Once the boundary is drawn, three transactions
cross it: matter, work, and heat. A system that exchanges all three
with its surroundings is an \engterm{open system}; a system that
exchanges work and heat but not matter is a \engterm{closed system}; a
system that exchanges nothing is an \engterm{isolated system}. The
distinction is set by the boundary, not by the contents.

A bank account is an accounting equation: deposits minus withdrawals
equals the change in balance. Thermodynamics is the same equation
applied to energy. The deposits are heat $Q$ added to the system and
work $W$ done on the system; the change in balance is the change
$\Delta E$ in the system's internal stored energy. The first law of
thermodynamics, in the form we will use throughout,
\[
\Delta E = Q + W,
\]
is the engineering accounting identity that every later chapter of
this volume builds on. The convention used here (work positive when
done on the system) is the one we will keep; some textbooks invert
the sign of $W$ \cite[ch.~2]{text:cengel-boles}. The chosen convention
is internally consistent; the reader who imports a result from another
text must check the sign on $W$ before using it.
Figure~\ref{fig:vol04ch01:first-law-signs} fixes the convention in a
single diagram.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-first-law-signs.tex}

The accounting equation is exact. It is also unfalsified across two
centuries of experiment, despite many proposals to extract net work
from no input. The exactness has consequences:

\begin{itemize}[leftmargin=*]
  \item Energy is conserved across every closed boundary. A balance
    that fails to close points to an unaccounted flow, not to a
    failure of the law.
  \item Energy in transit has two engineering modes: heat (transfer
    driven by temperature difference) and work (transfer driven by
    force-times-displacement, voltage-times-charge, pressure-times-volume,
    chemical-potential-times-mass, or any equivalent product). The
    book keeps the two modes separate because the second law treats
    them asymmetrically.
  \item Internal energy is the cumulative state of the system; heat
    and work are not properties of the system, they are properties of
    a process the system undergoes. We will return to this distinction
    in section~\ref{sec:state-vs-path}.
\end{itemize}

The accounting identity has the same form for any control volume in
any domain. A battery, a steam turbine, a cell, a city, and a planet
all admit energy balances. The art of thermodynamic analysis is
choosing the boundary that makes the balance illuminating.

\begin{archetype}[Balance]
A balance equation states that the time rate of change of a conserved
quantity inside a control volume equals the net inflow minus the net
outflow, plus internal generation. Mass, momentum, energy, charge, and
species each obey balance equations of this form. The thermodynamic
balance is energy, the second-law balance is entropy (with a
generation term that cannot be negative), and Volume V will add
species balances for reacting systems. The reader who can write the
balance for any flow across any boundary owns the working core of the
engineering analysis of energy.
\end{archetype}

\section{First law: closed and open systems}\pathtag{core}

The first law for a \engterm{closed system} undergoing a process from
state 1 to state 2 is
\[
\Delta U = Q + W,
\]
where $U$ is the internal energy (a state function), $Q$ the heat
crossing the boundary into the system, and $W$ the work done on the
system. For a closed system at rest with no change in kinetic or
potential energy, $\Delta E = \Delta U$. When kinetic or potential
energy changes are non-negligible (an aircraft on take-off, a falling
body in a calorimeter), the full $\Delta E = \Delta U + \Delta KE +
\Delta PE$ is used.

\engterm{Boundary work} for a quasi-static process in a closed system
of fluid at pressure $p$ and volume $V$ is
\[
W = -\int_{V_1}^{V_2} p \,dV.
\]
The sign reflects our convention: when the system expands ($dV > 0$),
the surroundings do negative work on the system; the system delivers
work to the surroundings. The integral is path-dependent: the work
done by a gas in expanding from $(p_1, V_1)$ to $(p_2, V_2)$ depends
on the trajectory through the $p\textnormal{-}V$ plane, not only on
the endpoints. The four canonical quasi-static processes have
closed-form work integrals:

\begin{itemize}[leftmargin=*]
  \item Isobaric ($p$ constant): $W = -p (V_2 - V_1)$.
  \item Isochoric ($V$ constant): $W = 0$.
  \item Isothermal ideal gas: $W = -nRT \ln(V_2/V_1)$.
  \item Adiabatic ideal gas ($pV^\gamma$ constant): $W = (p_2 V_2 -
    p_1 V_1)/(\gamma - 1)$.
\end{itemize}

The four processes are the building blocks of every cycle in chapter~3
of this volume. Figure~\ref{fig:vol04ch01:pv-processes} draws all four
leaving a common start state, so the difference in the work (the area
under each curve) is visible at a glance.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-pv-processes.tex}

The script \texttt{code/ideal\_gas\_processes.py} evaluates $W$, $Q$,
and $\Delta S$ for each of the four processes from the same input state
and reproduces the worked numbers of section~\ref{sec:state-vs-path}.

The first law for an \engterm{open system} (also called a control
volume) includes mass flow. For a control volume at steady state with
inlet stream at enthalpy $h_1$ and outlet stream at $h_2$, mass flow
rate $\dot{m}$, heat rate $\dot{Q}$ added to the control volume, and
shaft work rate $\dot{W}_s$ done on the control volume,
\[
0 = \dot{Q} + \dot{W}_s + \dot{m}\left(h_1 - h_2 + \tfrac{1}{2}(V_1^2 -
V_2^2) + g(z_1 - z_2)\right).
\]
Figure~\ref{fig:vol04ch01:control-volume} lays out the terms.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-control-volume.tex}

The \engterm{enthalpy} $h = u + pv$ (per unit mass) groups the
internal energy with the flow work $pv$ required to push mass across
the control volume's boundary. Whenever mass crosses a boundary at
non-zero pressure, the relevant energy property is enthalpy, not
internal energy. The kinetic and potential terms are usually small for
process equipment (turbines, compressors, heat exchangers) and large
for nozzles, diffusers, and aircraft engines. The script
\texttt{code/first\_law\_balance.py} evaluates the residual of this
balance for any chosen control volume and solves for shaft power; it
is verified on the steam turbine of the Calculation exercises.

\subsection{Deriving the control-volume energy equation}

The open-system balance was stated above; it earns trust when derived
from the closed-system law rather than asserted. The derivation reuses
the one tool already in hand, the first law for a fixed mass, applied
over a short interval to a control mass that is partly inside the
control volume and partly in transit across its boundaries.

Fix a control volume with one inlet and one outlet, and choose a
control mass that, at time $t$, consists of the matter inside the
control volume plus a small slug of mass $\delta m_i$ poised to enter
at the inlet. At time $t + \delta t$ the same control mass consists of
the matter inside the control volume plus a small slug $\delta m_e$ that
has just left through the outlet. The closed-system first law applied to
this control mass over the interval is
\[
\mathrm{d}E_{\text{cm}} = \delta Q + \delta W,
\]
where $\delta W$ collects every form of work crossing the control
mass's boundary. Two kinds of work appear. The first is the shaft work
$\delta W_s$ and any other non-flow work crossing the fixed boundary of
the control volume. The second is the work done to push the slugs across
the inlet and outlet, the \engterm{flow work} drawn in
Figure~\ref{fig:vol04ch01:flow-work}. The upstream fluid pushes the
inlet slug in against pressure $p_i$ through the slug's volume
$v_i\,\delta m_i$, doing work $+p_i v_i\,\delta m_i$ on the control mass;
the control mass pushes the outlet slug out against $p_e$, doing work
$-p_e v_e\,\delta m_e$ on the surroundings. So
\[
\delta W = \delta W_s + p_i v_i\,\delta m_i - p_e v_e\,\delta m_e.
\]

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-flow-work.tex}

The energy of the control mass changes for two reasons: the energy
stored inside the control volume changes by $\mathrm{d}E_{cv}$, and the
inlet slug carried in its own energy while the outlet slug carried some
out. Each slug carries internal, kinetic, and potential energy per unit
mass, written $e = u + \tfrac12 V^2 + gz$. Bookkeeping the control mass
between the two instants,
\[
\mathrm{d}E_{\text{cm}} = \mathrm{d}E_{cv} + e_e\,\delta m_e -
e_i\,\delta m_i.
\]
Substitute this and the work expression into the first law:
\[
\mathrm{d}E_{cv} + e_e\,\delta m_e - e_i\,\delta m_i = \delta Q +
\delta W_s + p_i v_i\,\delta m_i - p_e v_e\,\delta m_e.
\]
Collect the inlet terms and the outlet terms, and the flow work folds
into each stream's energy:
\[
\mathrm{d}E_{cv} = \delta Q + \delta W_s +
(e_i + p_i v_i)\,\delta m_i - (e_e + p_e v_e)\,\delta m_e.
\]
The grouping $e + pv = u + pv + \tfrac12 V^2 + gz = h + \tfrac12 V^2 +
gz$ is exactly the enthalpy plus the mechanical terms. Dividing by
$\delta t$ and passing to rates gives the general control-volume energy
equation,
\[
\frac{\mathrm{d}E_{cv}}{\mathrm{d}t} = \dot{Q} + \dot{W}_s +
\dot{m}_i\!\left(h_i + \tfrac12 V_i^2 + gz_i\right) -
\dot{m}_e\!\left(h_e + \tfrac12 V_e^2 + gz_e\right).
\]
At steady state the stored energy is constant, $\mathrm{d}E_{cv}/
\mathrm{d}t = 0$, and a single stream conserves mass, $\dot{m}_i =
\dot{m}_e = \dot{m}$, which returns the steady-flow balance used above.
The enthalpy was not assumed; it arose because the flow work $pv$
attaches to every stream that crosses a boundary at non-zero pressure.
That is the content of the claim that open systems are accounted in
enthalpy and closed systems in internal energy: the difference is the
flow work, and the derivation locates it precisely.

\engterm{Specific heats}. For a closed system at constant volume,
$dU = c_v(T) \,dT$ per unit mass, with $c_v$ the specific heat at
constant volume. For a closed system at constant pressure, $dH =
c_p(T) \,dT$, with $c_p$ the specific heat at constant pressure. The
two satisfy $c_p - c_v = R$ for an ideal gas (per mole) or $c_p -
c_v = R/M$ per unit mass with $M$ the molar mass. The ratio $\gamma =
c_p/c_v$ enters adiabatic processes and the speed of sound and is the
single most useful dimensionless property of a gas: $\gamma \approx
1.4$ for diatomic air at room temperature, $\gamma \approx 1.67$ for
monatomic argon, $\gamma \approx 1.3$ for steam at moderate
conditions.

\begin{principle}[The first law is an accounting identity]
The change in the energy of any thermodynamic system equals the net
energy crossing its boundary as heat, work, and (for open systems)
mass flow. The identity is exact, has units of joules (per unit time
for the rate form), and applies to any control volume in any domain.
Every later analysis of a thermal, electrical, chemical, or mechanical
process is the first law evaluated on a chosen boundary, with the
chosen work and heat terms identified. A balance that does not close
is a measurement or modelling error, not a violation of the law.
\end{principle}

\section{Second law: entropy and direction}\pathtag{core}

The first law says that energy is conserved. It does not say which way
energy is allowed to move. A flask of hot tea at 80~\degC{} placed in a
room at 20~\degC{} cools to 20~\degC{}; the room does not spontaneously
warm the tea to 80~\degC{}, even though both directions would conserve
energy. The asymmetry is the second law.

The two classical statements of the second law:

\begin{itemize}[leftmargin=*]
  \item \engterm{Clausius statement}: heat cannot, of itself, pass
    from a colder body to a hotter body. ``Of itself'' means without a
    compensating change elsewhere; refrigerators move heat from cold
    to hot, but only at the cost of input work that ends up as heat
    rejected at the hot side. The statement is the canonical
    expression of the directionality of heat flow
    \cite{hist:clausius1850}.
  \item \engterm{Kelvin-Planck statement}: no cyclic device can take
    heat from a single reservoir and convert it entirely to work. A
    cyclic heat engine must reject some heat to a colder reservoir;
    the ratio of work extracted to heat input is bounded by Carnot's
    theorem (chapter~3 of this volume).
\end{itemize}

The Kelvin-Planck statement is the schematic of
Figure~\ref{fig:vol04ch01:heat-engine}: a cyclic engine must reject
$Q_c$ to a cold reservoir, and the work it delivers is bounded by the
Carnot value.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-heat-engine.tex}

The two statements are equivalent: a device that violated one could be
combined with a Carnot engine or refrigerator to produce a device that
violated the other. The proof is in Carnot's 1824 treatise in
embryonic form \cite{hist:carnot1824} and was made formal by Clausius
in 1850 \cite{hist:clausius1850}.

The bookkeeping form of the second law uses \engterm{entropy}, the
state function $S$ with
\[
dS = \frac{\delta Q_{\text{rev}}}{T},
\]
where $\delta Q_{\text{rev}}$ is the heat transferred to the system
along a reversible path and $T$ is the system's absolute temperature.
Entropy has units of joules per kelvin. For a closed system
undergoing any process between two states,
\[
\Delta S \geq \int \frac{\delta Q}{T},
\]
with equality for a reversible process and strict inequality for
irreversible processes. For an isolated system ($\delta Q = 0$), the
entropy can only increase or stay constant:
\[
\Delta S_{\text{isolated}} \geq 0.
\]
This is the form most often cited as the second law: the entropy of
the universe (taken as an isolated system) does not decrease.

The second law in rate form for an open system is the \engterm{entropy
balance}:
\[
\frac{dS_{cv}}{dt} = \sum \frac{\dot{Q}_j}{T_j} + \sum \dot{m}_i s_i -
\sum \dot{m}_e s_e + \dot{S}_{\text{gen}},
\]
where $T_j$ is the temperature at the boundary across which heat
$\dot{Q}_j$ enters, $s_i$ and $s_e$ are the inlet and exit specific
entropies, and $\dot{S}_{\text{gen}} \geq 0$ is the
\engterm{entropy-generation rate} inside the control volume. The
inequality $\dot{S}_{\text{gen}} \geq 0$ is the second law's
quantitative content for open systems: any process generates entropy
inside the control volume in an amount that depends on the process's
irreversibilities (heat transfer across a finite temperature
difference, friction, unrestrained expansion, mixing, chemical
reaction at finite affinity). Entropy generation is the engineering
diagnostic of loss; a process with $\dot{S}_{\text{gen}} = 0$ is
reversible and extracts the maximum possible work from its available
inputs. Real processes have $\dot{S}_{\text{gen}} > 0$ and extract
less.

The reader who finishes this chapter knows the second law as three
things at once: a direction-of-spontaneous-change rule, a bound on the
efficiency of heat engines, and an entropy balance with a
non-negative generation term. Chapter~4 of this volume provides the
microscopic interpretation that makes the entropy balance more than
bookkeeping.

\begin{mastery}
The bound $\dot{S}_{\text{gen}} \geq 0$ converts into a bound on lost
work through the \engterm{Gouy-Stodola theorem}. For a process drawing
on an environment at temperature $T_0$, the rate of lost work, the
shortfall between the actual work and the reversible best, is
\[
\dot{W}_{\text{lost}} = T_0 \, \dot{S}_{\text{gen}}.
\]
Every kelvin of entropy generated per second at a process with a
$300\,\si{\kelvin}$ environment destroys $300\,\text{W}$ of work
potential. This is the working content of second-law (exergy)
analysis: the engineer ranks the components of a plant by their
$T_0 \dot{S}_{\text{gen}}$ and attacks the largest term first. A heat
exchanger transferring $\dot{Q}$ across a temperature drop from $T_1$
to $T_2$ generates $\dot{S}_{\text{gen}} = \dot{Q}(1/T_2 - 1/T_1)$, so
the lost work scales with the temperature drop. Volume V develops the
exergy accounting; the present identity is the bridge from the entropy
balance to a euro figure.
\end{mastery}

\section{Third law}\pathtag{standard}

The third law of thermodynamics, in the form due to Walther Nernst
(1906) and made precise by Planck (1911), states that the entropy of a
pure crystalline substance at absolute zero is zero. The law has two
working consequences.

The first is the impossibility of reaching absolute zero in a finite
number of operations. Each cooling stage in a refrigeration cascade
extracts a finite amount of entropy; as $T \to 0$, the amount of
entropy that can be extracted per cycle approaches zero, and the
number of cycles required to reach $T = 0$ diverges. The lowest
temperatures achieved in laboratory experiments (nanokelvin in
laser-cooled atomic gases, picokelvin in nuclear-spin systems) are
finite for this reason.

The second is the use of absolute entropies. With the third law as
zero, the entropy of any substance at temperature $T$ above absolute
zero is obtained by integrating $c_p(T)/T \,dT$ from $0$ to $T$, plus
the latent heats divided by transition temperatures at any phase
transitions. Standard chemical and physical reference tables list
absolute molar entropies $S^\circ$ at the standard state ($25\,\degC$,
$1\,\text{atm}$) for thousands of substances. The absolute entropies
allow direct calculation of $\Delta S$ for chemical reactions from
tabulated values rather than from path integration.
Figure~\ref{fig:vol04ch01:entropy-temperature} shows the construction:
the curve climbs from the third-law zero by accumulating $c_p/T$, with
a vertical jump at each phase transition. A short table of standard
molar entropies is provided in \texttt{data/standard\_entropy.csv}.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-entropy-temperature.tex}

The third law is the least operationally useful of the three for
engineering practice at ordinary temperatures, but it is the law that
makes entropy a numerical quantity rather than a difference. In low-
temperature engineering (superconducting magnets, cryogenic liquids,
quantum computing hardware), the third law is the governing
constraint.

\subsection{Computing an absolute entropy, and why the integral converges}

The claim that the entropy at temperature $T$ is the integral of $c_p/T$ from
absolute zero raises an immediate worry: the integrand carries a $1/T$ that blows
up as $T \to 0$, so the integral would seem to diverge and no finite absolute
entropy would exist. The third law resolves the worry by forcing the numerator to
vanish faster than the denominator. A stable substance must have its heat capacity
go to zero at absolute zero, $c_p \to 0$ as $T \to 0$, and for a crystalline solid
the low-temperature behaviour is the \engterm{Debye law} $c_p \approx a T^3$, with
$a$ a constant set by the material's speed of sound. The entropy integral near the
origin is then
\[
S(T) = \int_0^T \frac{c_p(T')}{T'}\,\mathrm{d}T' \approx
\int_0^T \frac{a T'^3}{T'}\,\mathrm{d}T' = \int_0^T a T'^2\,\mathrm{d}T'
= \frac{a T^3}{3},
\]
which is finite and, more, equals $c_p(T)/3$ at the same temperature: the entropy
and the heat capacity vanish together as $T^3$. The $1/T$ singularity is defeated
because the vibrational modes of the crystal freeze out as the temperature falls,
each mode ceasing to absorb heat once $kT$ drops below its quantum of energy, so
there is simply no heat capacity left to integrate. The third law is the statement
that this freezing is complete: at $T = 0$ every mode is in its ground state, the
crystal has one accessible microstate, and the entropy floor is zero.

The full absolute molar entropy of a substance at a standard temperature is then a
sum of a convergent integral over each phase plus a latent-heat jump at each
transition,
\[
S^\circ(T) = \int_0^{T_m}\!\frac{c_p^{\text{solid}}}{T'}\,\mathrm{d}T'
+ \frac{\Delta H_{\text{fus}}}{T_m}
+ \int_{T_m}^{T_b}\!\frac{c_p^{\text{liquid}}}{T'}\,\mathrm{d}T'
+ \frac{\Delta H_{\text{vap}}}{T_b}
+ \int_{T_b}^{T}\!\frac{c_p^{\text{gas}}}{T'}\,\mathrm{d}T',
\]
the construction Figure~\ref{fig:vol04ch01:entropy-temperature} draws as a curve
climbing from the third-law zero with a vertical step at each phase change. Every
term is measurable: the heat capacities by calorimetry, the latent heats at the
transition temperatures. The third law supplies the lower limit of the first
integral, without which the whole sum would float by an unknown constant and only
entropy differences would be defined. This is why a chemist can read a single
number $S^\circ$ from a table and compute the entropy change of a reaction as
$\Delta S^\circ = \sum \nu_i S^\circ_i$ over products minus reactants, the same way
enthalpies of formation are summed, a convenience that rests entirely on the
common $S = 0$ floor the third law fixes.

The floor holds only for a \engterm{perfect} crystal. A solid frozen with disorder
locked in, a glass, a mixed crystal, or a molecule that can sit two ways in the
lattice, retains a \engterm{residual entropy} at absolute zero equal to
$k\ln \Omega$ with $\Omega$ the number of frozen-in configurations. Carbon monoxide
is the textbook case: the molecule is nearly symmetric, so each one freezes head-to-tail
or tail-to-head almost at random, giving two orientations per molecule and a
predicted residual entropy of $R\ln 2 = 5.76\,\text{J}/(\text{mol}\cdot\si{\kelvin})$
per mole, close to the $4.6$ measured as the gap between the calorimetric entropy
integrated from zero and the spectroscopic entropy computed from the gas's energy
levels. The residual entropy is not a violation of the third law; it is the
signature of a crystal that never reached its single true ground state because the
molecules froze in place before they could order, the barrier to reorientation
having grown faster than thermal energy fell. The engineering reading is that the
$S = 0$ floor is a property of equilibrium order, and a real material carries to
absolute zero whatever disorder its cooling rate trapped, a caution that recurs in
the annealing of metals and the setting of glasses.

\begin{mastery}
The unreachability of absolute zero is quantified by adiabatic
demagnetisation, the workhorse of sub-kelvin cooling. A paramagnetic
salt is magnetised at a starting temperature while in thermal contact
with a bath, dumping its spin entropy; the bath is then removed and the
field lowered adiabatically (at constant entropy), which forces the
temperature down as the spins re-randomise. Each stage cools by a
factor set by the ratio of final to initial field, but the entropy
available to extract shrinks as $T \to 0$ because the spin system
freezes into its ground state, so the gain per stage collapses and no
finite sequence reaches $T = 0$. This is the operational content of the
Nernst statement: the isentropes of any substance converge as
$T \to 0$, and a process at constant entropy cannot bridge the gap to
the $T = 0$ isotherm in a finite number of steps. The same convergence
is why standard molar entropies are well defined: all substances share
the $S = 0$ floor.
\end{mastery}

\section{State functions vs path functions}\pathtag{core}
\label{sec:state-vs-path}

A \engterm{state function} is a property of the system that depends
only on the system's current state, not on how the system arrived
there. Internal energy $U$, enthalpy $H$, entropy $S$, Helmholtz free
energy $A$, Gibbs free energy $G$, temperature $T$, pressure $p$,
volume $V$, density $\rho$, and chemical composition are all state
functions. Their differentials are exact: $dU$, $dH$, $dS$ all
integrate around any closed cycle to zero.

A \engterm{path function} is a quantity associated with a process, not
with a state. Heat $Q$ and work $W$ are the canonical path functions:
the heat absorbed and the work done in going from state 1 to state 2
depend on the path through the $p$-$V$ plane (or any equivalent state
plane), not only on the endpoints. The differentials are inexact:
$\delta Q$ and $\delta W$ do not integrate to a state function around
a cycle. A heat engine extracts net work from a cyclic process
precisely because $W$ and $Q$ are path functions; $\oint dU = 0$
around any cycle, but $\oint \delta W \neq 0$ in general.

The distinction is the source of much of thermodynamics' analytical
power. The state-function property allows a calculation: to find
$\Delta H$ for a process, find any convenient reversible path between
the same endpoints and integrate. The path-function property forces a
real calculation: to find $W$ or $Q$, the actual process must be
modelled or measured.
Figure~\ref{fig:vol04ch01:state-vs-path} draws two paths between one
pair of states: the change in every state function is the same for
both, while the work (the area under each path) differs.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-state-vs-path.tex}

\engterm{Common state-function identities for an ideal gas}. For an
ideal gas:
\begin{align*}
\Delta U &= n c_v \Delta T, \\
\Delta H &= n c_p \Delta T, \\
\Delta S &= n c_v \ln(T_2/T_1) + nR \ln(V_2/V_1), \\
\Delta S &= n c_p \ln(T_2/T_1) - nR \ln(p_2/p_1).
\end{align*}
The last two are the same equation in two parameter systems; they
follow from $dU = T\,dS - p\,dV$ (the combined first and second laws)
applied to an ideal gas with constant $c_v$. The reader who memorises
one form should be able to derive the other in under a minute.

\engterm{Cycle integrals}. Around a closed cycle the state functions
return to their initial values, so $\oint dU = 0$, $\oint dH = 0$,
$\oint dS = 0$. The first-law cycle integral $\oint \delta Q = -\oint
\delta W$ relates the net heat absorbed to the net work done; the
ratio (work to heat input) is the cycle's thermal efficiency, and the
Carnot bound (chapter~3) is the maximum achievable for given hot and
cold reservoir temperatures.

\subsection{Exact and inexact differentials: the calculus behind the
distinction}

The words state function and path function have a precise calculus behind
them, and the engineer who carries the calculus rather than the slogan can
predict which combinations of measured quantities will integrate to a
property and which will not. A quantity expressed as a differential form in
two independent variables, say
\[
\mathrm{d}f = M(x,y)\,\mathrm{d}x + N(x,y)\,\mathrm{d}y,
\]
is an \engterm{exact differential} when there exists a function $f(x,y)$
whose partial derivatives are $M$ and $N$, that is, $M = (\partial f/
\partial x)_y$ and $N = (\partial f/\partial y)_x$. The test is the equality
of the mixed second partials of $f$, which any twice-differentiable function
obeys:
\[
\left(\frac{\partial M}{\partial y}\right)_x =
\left(\frac{\partial N}{\partial x}\right)_y.
\]
When this \engterm{Euler reciprocity} condition holds, the integral of
$\mathrm{d}f$ between two points is $f$ at the endpoints minus $f$ at the
start, independent of the path, and the integral around any closed loop is
zero. When it fails, no such $f$ exists, the integral depends on the path
taken, and the form is an \engterm{inexact differential}, written here with
$\delta$ rather than $\mathrm{d}$.

Apply the test to the boundary work of an ideal gas. With $p = nRT/V$, the
work form is $\delta W = -p\,\mathrm{d}V = -(nRT/V)\,\mathrm{d}V$, which in
the variables $(T, V)$ reads $\delta W = 0 \cdot \mathrm{d}T -
(nRT/V)\,\mathrm{d}V$, so $M = 0$ and $N = -nRT/V$. The reciprocity test
compares $(\partial M/\partial V)_T = 0$ with $(\partial N/\partial T)_V =
-nR/V$, and the two are not equal: $\delta W$ fails the test and is inexact,
which is the calculus statement of the path dependence drawn in
Figure~\ref{fig:vol04ch01:inexact-differential}. The heat $\delta Q =
\mathrm{d}U - \delta W = nc_v\,\mathrm{d}T + (nRT/V)\,\mathrm{d}V$ has $M =
nc_v$ and $N = nRT/V$, and the test compares $(\partial M/\partial V)_T = 0$
with $(\partial N/\partial T)_V = nR/V$: $\delta Q$ is inexact too.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-inexact-differential.tex}

The second law performs a small miracle on the inexact heat form. Divide
$\delta Q$ by the temperature $T$, the \engterm{integrating factor} the
second law supplies:
\[
\frac{\delta Q}{T} = \frac{nc_v}{T}\,\mathrm{d}T + \frac{nR}{V}\,\mathrm{d}V.
\]
Now $M = nc_v/T$ and $N = nR/V$, and the reciprocity test compares
$(\partial M/\partial V)_T = 0$ with $(\partial N/\partial T)_V = 0$: the two
agree, so $\delta Q/T$ is exact. The function whose differential it is, is
the entropy, and integrating recovers $\Delta S = nc_v\ln(T_2/T_1) +
nR\ln(V_2/V_1)$, the form derived earlier. The deep content of the second
law, read through the calculus, is that an integrating factor exists that
turns the inexact heat form into the exact differential of a state function,
and that the factor is the absolute temperature for every substance. The
first law gives energy as the state function behind the work-plus-heat sum;
the second law gives entropy as the state function behind the heat-over-
temperature ratio. Both laws are existence statements about functions hiding
behind inexact forms.

\subsection{The cyclic and reciprocal relations}

A state described by three variables linked by one equation of state, such
as $p$, $V$, $T$ tied by $f(p,V,T) = 0$, has only two of the three free at
once, and the partial derivatives among them obey two identities that recur
throughout thermodynamics. Treat any one variable as a function of the other
two, say $V = V(T, p)$, and write its differential,
\[
\mathrm{d}V = \left(\frac{\partial V}{\partial T}\right)_p \mathrm{d}T +
\left(\frac{\partial V}{\partial p}\right)_T \mathrm{d}p.
\]
The \engterm{reciprocal relation} follows by holding one variable fixed: at
constant $p$, dividing through by $\mathrm{d}T$ gives nothing new, but
inverting the roles of $V$ and $T$ shows that
\[
\left(\frac{\partial V}{\partial p}\right)_T =
\frac{1}{(\partial p/\partial V)_T},
\]
each derivative the reciprocal of its inverted partner taken at the same
fixed variable. The \engterm{cyclic relation}, also called the triple-product
rule, is the one that surprises the newcomer. Set $\mathrm{d}V = 0$ in the
differential above, which fixes $V$ and forces a relation between the
remaining changes $\mathrm{d}T$ and $\mathrm{d}p$:
\[
0 = \left(\frac{\partial V}{\partial T}\right)_p \mathrm{d}T +
\left(\frac{\partial V}{\partial p}\right)_T \mathrm{d}p
\quad\Longrightarrow\quad
\left(\frac{\partial p}{\partial T}\right)_V =
-\frac{(\partial V/\partial T)_p}{(\partial V/\partial p)_T}.
\]
Rearranging into a symmetric product gives the cyclic relation in its
memorable form,
\[
\left(\frac{\partial p}{\partial T}\right)_V
\left(\frac{\partial T}{\partial V}\right)_p
\left(\frac{\partial V}{\partial p}\right)_T = -1,
\]
where the minus sign, not the plus a careless chain-rule cancellation would
predict, is the whole point. The three derivatives chase the variables
around a closed loop, and the constraint $f(p,V,T) = 0$ inserts the negative
sign. The relation earns its keep by trading a derivative that is hard to
measure for two that are easy: the pressure rise of a sealed rigid vessel
heated by one kelvin, $(\partial p/\partial T)_V$, is awkward to measure
directly but follows at once from the thermal expansion coefficient $\alpha =
(1/V)(\partial V/\partial T)_p$ and the isothermal compressibility $\kappa_T
= -(1/V)(\partial V/\partial p)_T$, both routine measurements, as
$(\partial p/\partial T)_V = \alpha/\kappa_T$. A sealed glass thermometer
bulb or a rigid pressure vessel left in the sun builds pressure at exactly
this rate, and the cyclic relation is how the designer sizes the relief for
it without ever heating a sealed vessel to read the slope.

\subsection{The Maxwell relations from the potentials}

The four common energy potentials each furnish a Maxwell relation by the
same reciprocity test that classified the differential forms above. The
construction is mechanical once the differential of each potential is in
hand. Internal energy carries $\mathrm{d}U = T\,\mathrm{d}S - p\,\mathrm{d}V$
(at fixed composition), so reading off $T = (\partial U/\partial S)_V$ and
$-p = (\partial U/\partial V)_S$ and equating the mixed second partials of
$U$ gives
\[
\left(\frac{\partial T}{\partial V}\right)_S =
-\left(\frac{\partial p}{\partial S}\right)_V.
\]
Helmholtz energy carries $\mathrm{d}A = -S\,\mathrm{d}T - p\,\mathrm{d}V$,
and the same step gives
$(\partial S/\partial V)_T = (\partial p/\partial T)_V$. Gibbs energy carries
$\mathrm{d}G = -S\,\mathrm{d}T + V\,\mathrm{d}p$, giving
$(\partial S/\partial p)_T = -(\partial V/\partial T)_p$. Enthalpy carries
$\mathrm{d}H = T\,\mathrm{d}S + V\,\mathrm{d}p$, giving $(\partial T/\partial
p)_S = (\partial V/\partial S)_p$. The four relations share a pattern that
the \engterm{thermodynamic square} of
Figure~\ref{fig:vol04ch01:maxwell-square} encodes: the natural variables sit
at the corners, the potentials on the edges between their pairs, and a
Maxwell relation is read by walking the corners. The working value of the
set is uniform. Each Maxwell relation equates a derivative involving entropy,
which no instrument reads directly, to a derivative of the mechanical
variables $p$, $V$, $T$, which any well-found laboratory measures. The
relation $(\partial S/\partial V)_T = (\partial p/\partial T)_V$, for one,
lets the entropy rise of an isothermal expansion be computed from the
pressure-temperature slope of the substance, and through it the heat absorbed
in that expansion, all from $pVT$ data and never from a calorimeter run at
the expansion conditions.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-maxwell-square.tex}

A short worked use closes the construction. The internal energy of a real
fluid changes with volume even at constant temperature, unlike an ideal gas,
and the size of that change is exactly what a Maxwell relation delivers.
Starting from $\mathrm{d}U = T\,\mathrm{d}S - p\,\mathrm{d}V$ and dividing by
$\mathrm{d}V$ at constant $T$,
\[
\left(\frac{\partial U}{\partial V}\right)_T =
T\left(\frac{\partial S}{\partial V}\right)_T - p
= T\left(\frac{\partial p}{\partial T}\right)_V - p,
\]
where the Maxwell relation from Helmholtz energy replaced the entropy
derivative in the second step. This is the \engterm{thermodynamic equation of
state}, and it carries an immediate test for ideality. For an ideal gas
$p = nRT/V$ gives $(\partial p/\partial T)_V = nR/V = p/T$, so the bracket
$T(\partial p/\partial T)_V - p = p - p = 0$: the internal energy of an ideal
gas does not depend on volume at fixed temperature, the property that let
every ideal-gas example of this chapter write $\Delta U = nc_v\Delta T$ with
no volume term. A real gas with attractive forces has $T(\partial p/\partial
T)_V - p > 0$, so its internal energy rises as it expands at constant
temperature, the energy going into pulling molecules apart against their
attraction, and that positive $(\partial U/\partial V)_T$ is what makes the
Joule-Thomson coefficient of the throttling example non-zero.

\section{Internal energy, enthalpy, free energies}\pathtag{standard}

Five state functions of energy underwrite the bulk of engineering
thermodynamic analysis. Each is defined by a combination of other
state functions, and each is the natural variable for a particular
process geometry.

\engterm{Internal energy} $U$. The total energy stored microscopically
in the system: kinetic energy of molecules, potential energy of
intermolecular forces, electronic energy, and the chemical-bond
energies. The natural variables of $U$ are $S$ and $V$:
\[
dU = T \,dS - p \,dV + \sum_i \mu_i \,dN_i,
\]
where $\mu_i$ is the chemical potential of species $i$ and $N_i$ the
number of moles. The chemical-potential term enters when the system's
composition can change (mass transfer, chemical reaction, phase
change).

\engterm{Enthalpy} $H = U + pV$. The natural variables of $H$ are $S$
and $p$:
\[
dH = T \,dS + V \,dp + \sum_i \mu_i \,dN_i.
\]
For a constant-pressure process, $\Delta H = Q$: enthalpy is the heat
exchanged by the system in any constant-pressure process. Combustion
processes, evaporation, and most chemical reactions in open laboratory
or industrial vessels are constant-pressure, so enthalpy is the
working quantity in chemical thermodynamics. Flow processes (where the
system is open and mass crosses a boundary) likewise use enthalpy
rather than internal energy because of the $pv$ flow-work term.

\engterm{Helmholtz free energy} $A = U - TS$. The natural variables of
$A$ are $T$ and $V$:
\[
dA = -S \,dT - p \,dV + \sum_i \mu_i \,dN_i.
\]
For a closed system at constant temperature and volume, $\Delta A$ is
the maximum useful work that can be extracted from a state change.
Constant-$T$ constant-$V$ processes are rare in engineering (the
typical closed-vessel reaction is at constant volume but variable
temperature), but $A$ is the natural function for statistical
mechanics: the canonical-ensemble partition function gives $A$
directly (chapter~4 of this volume).

\engterm{Gibbs free energy} $G = H - TS = U + pV - TS$. The natural
variables of $G$ are $T$ and $p$:
\[
dG = -S \,dT + V \,dp + \sum_i \mu_i \,dN_i.
\]
For a closed system at constant temperature and pressure, $\Delta G$
is the maximum useful work (other than $pV$ work against the
atmosphere) that can be extracted. Most chemical reactions of
engineering interest happen at approximately constant temperature and
pressure (open vessels at atmospheric, biological systems at body
temperature), so $G$ is the working free energy of chemistry and
biology. The equilibrium condition for a reaction at constant $T$, $p$
is $\Delta G = 0$; the criterion for spontaneity is $\Delta G < 0$;
the criterion for non-spontaneity is $\Delta G > 0$. Volume V develops
this machinery; Volume VI applies it to bioenergetics.

\engterm{Grand potential} $\Omega = U - TS - \sum_i \mu_i N_i$. The
natural variables are $T$, $V$, and $\mu_i$. The grand potential is
the working function for open-system statistical mechanics and for
problems where chemical potential rather than composition is the
externally controlled variable (membranes, electrochemistry,
semiconductor doping).

The five free energies are related by Legendre transforms; the
\engterm{Maxwell relations} that follow from the equality of cross
partial derivatives connect them and provide identities that are
indispensable in real-gas and phase-change calculations. Standard
examples:
\[
\left(\pdv{S}{V}\right)_T = \left(\pdv{p}{T}\right)_V, \quad
\left(\pdv{S}{p}\right)_T = -\left(\pdv{V}{T}\right)_p.
\]
The Maxwell relations let the engineer trade an experimentally
inaccessible derivative for one obtainable from $pVT$ data. Chapter~2
of this volume uses them to compute departures from ideal-gas
behaviour.

\section{Worked examples: gas compression, mixing, phase change}\pathtag{standard}

The state-function machinery becomes practical through repeated
application to canonical processes. Three examples cover the working
core.

\subsection{Isothermal and adiabatic compression of an ideal gas}

A piston compresses one mole of air (an ideal gas with $\gamma = 1.4$,
$R = 8.314\,\text{J}/(\text{mol}\cdot\text{K})$, $c_v = 5R/2$) from
$p_1 = 1\,\text{atm}$, $V_1 = 24\,\si{\liter}$ to half its initial
volume. The compression can be done isothermally (held at $T_1 =
293\,\si{\kelvin}$ by contact with a reservoir) or adiabatically
(quickly, with no heat transfer).

For the isothermal compression, $T$ is constant, so $\Delta U = 0$ and
$Q = -W$. The work done on the gas is $W = -nRT \ln(V_2/V_1) = -1
\times 8.314 \times 293 \times \ln(0.5) = +1690\,\text{J}$ (positive
because work is done on the gas; the convention is preserved). The
heat rejected to the reservoir is $-1690\,\text{J}$. The entropy
change of the gas is $\Delta S = nR \ln(V_2/V_1) = -5.76\,\text{J}/
\text{K}$, the reservoir's entropy change is $+5.76\,\text{J}/\text{K}$,
and the total entropy change is zero (a reversible isothermal
compression).

For the adiabatic compression, $Q = 0$ and $\Delta U = W$. The
end-state pressure and temperature follow from $pV^\gamma$ constant
and the ideal-gas law: $p_2 = p_1 (V_1/V_2)^\gamma = 2^{1.4} \times
1\,\text{atm} \approx 2.64\,\text{atm}$, $T_2 = T_1
(V_1/V_2)^{\gamma-1} = 2^{0.4} \times 293 \approx 387\,\si{\kelvin}$.
The work done on the gas is $W = n c_v \Delta T = 1 \times (5/2)
\times 8.314 \times (387 - 293) \approx 1960\,\text{J}$. The entropy
change of the gas is zero (reversible adiabatic), and the surroundings
exchange no heat, so the total entropy change is again zero.

The adiabatic compression takes more work than the isothermal because
the gas heats up; the increased pressure during compression demands
more work to push the piston the same distance. The same heating is
what makes a diesel engine's compression stroke ignite the fuel
charge: $T$ rises through the combustion temperature without an
electrical spark.

\subsection{Mixing of two ideal gases}

Two ideal gases at the same temperature and pressure, separated by a
partition, mix when the partition is removed. The internal energy of
the mixture equals the sum of the internal energies of the components
(an ideal-gas property), so $\Delta U = 0$ and $Q = 0$. No work is done
(the gases expand against each other, not against a piston), so $W =
0$ and the first-law balance is trivially $0 = 0 + 0$.

The entropy change is not zero. Each species expands into the full
volume; its partial pressure drops from the total pressure $p$ to its
partial pressure $x_i p$, where $x_i$ is its mole fraction. The
entropy change per mole of species $i$ is $-R \ln x_i$. For a mixture
of $n_A$ moles of $A$ and $n_B$ moles of $B$,
\[
\Delta S_{\text{mix}} = -R(n_A \ln x_A + n_B \ln x_B) > 0.
\]
The mixing entropy is positive because each $x_i < 1$ and the
logarithms are negative. The process is irreversible: separating the
two gases back into their original compartments requires work
(equivalent, in the ideal-gas limit, to the work of an isothermal
compression back to the original partial pressures). The script
\texttt{code/entropy\_of\_mixing.py} evaluates the formula and runs a
particle simulation whose bin-population entropy climbs to $R \ln 2$
per mole as an equimolar mixture equilibrates.

The mixing entropy is the engineering content of the second law for
separations. Distillation, gas absorption, membrane separation, and
crystallisation all undo a mixing-entropy increase, which costs work
in proportion to the entropy of the original mixing. Chapter~7 of
this volume develops the separation-process side; the present
calculation establishes the lower bound.

\subsection{Vapourisation of water}

A litre of water at $100\,\degC$ and $1\,\text{atm}$ is vapourised at
constant pressure. The latent heat of vapourisation of water at
$100\,\degC$ is $h_{fg} = 2257\,\text{kJ}/\text{kg}$ \cite[Table A-4]{
text:cengel-boles}. The water mass is $1\,\text{kg}$ (density
$1000\,\text{kg}/\text{m}^3$), so $Q = m h_{fg} = 2257\,\text{kJ}$.

The work done by the vapour against the atmosphere is $W = -p \Delta V
= -p(V_g - V_f)$. At $100\,\degC$ and $1\,\text{atm}$, the specific
volume of saturated vapour is $v_g \approx 1.673\,\text{m}^3/
\text{kg}$ and of saturated liquid $v_f \approx 1.04 \times
10^{-3}\,\text{m}^3/\text{kg}$, so $\Delta V \approx 1.67\,\text{m}^3
/\text{kg}$. The work is $W = -10^5 \,\text{Pa} \times 1.67\,
\text{m}^3 = -167\,\text{kJ}$ (negative: the system does work on the
atmosphere). The change in internal energy is $\Delta U = Q + W = 2257
- 167 = 2090\,\text{kJ}$.

The entropy change is $\Delta S = h_{fg}/T = 2257\,\text{kJ}/373\,
\si{\kelvin} = 6.05\,\text{kJ}/\text{K}$. The corresponding entropy
change of the surroundings (which gave up $2257\,\text{kJ}$ of heat at
the same temperature, idealised as reversible) is $-6.05\,\text{kJ}/
\text{K}$; the total entropy change is zero. A real kettle, of course,
delivers heat at a much higher temperature than the water's
$100\,\degC$; the actual surroundings entropy change is smaller in
magnitude, and the total entropy change is positive (irreversibility
due to finite-temperature-difference heat transfer). The saturation
properties used here are tabulated in
\texttt{data/water\_saturation.csv}.

\subsection{The Carnot cycle as an upper bound}

A cycle made of two isothermal and two reversible adiabatic segments,
the Carnot cycle, sets the efficiency ceiling for any heat engine
working between two reservoir temperatures. Take one mole of an ideal
gas with $\gamma = 1.4$ cycling between $T_h = 600\,\si{\kelvin}$ and
$T_c = 300\,\si{\kelvin}$, with an isothermal expansion that doubles
the volume at $T_h$. The heat absorbed in the high-temperature
expansion is $Q_h = nRT_h \ln 2 = 1 \times 8.314 \times 600 \times
0.693 = 3458\,\text{J}$. By the geometry of the cycle the entropy
delivered at the top equals the entropy removed at the bottom, so the
heat rejected is $Q_c = nRT_c \ln 2 = 1729\,\text{J}$, exactly half of
$Q_h$. The net work delivered is $Q_h - Q_c = 1729\,\text{J}$, and the
efficiency is
\[
\eta = \frac{W}{Q_h} = 1 - \frac{Q_c}{Q_h} = 1 - \frac{T_c}{T_h} =
1 - \frac{300}{600} = 0.50.
\]
The two heat exchanges happen at the single temperatures $T_h$ and
$T_c$, which is why the entropy transferred in and out are equal and
the efficiency reaches the bound. Any real engine between these two
reservoirs does worse, because real heat exchange happens across a
finite temperature difference and generates entropy.
Figures~\ref{fig:vol04ch01:carnot-pv} and~\ref{fig:vol04ch01:carnot-ts}
draw the cycle on $p$-$V$ and $T$-$S$ axes; on the second it is a
rectangle whose area is the net work. The script
\texttt{code/carnot\_cycle.py} computes the state points and confirms
that the cycle efficiency equals $1 - T_c/T_h$ to machine precision.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-carnot-pv.tex}

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-carnot-ts.tex}

\begin{estimation}
\textbf{How much electrical work does a household refrigerator consume
per day to maintain $4\,\degC$ inside a typical kitchen at
$22\,\degC$?}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A working estimate: the refrigerator must reject the heat it pumps out
plus the work input to its compressor. The heat to be pumped is the
sum of conductive heat leak through the cabinet walls, infiltration
when the door opens, and the heat of warm items placed inside.

The cabinet has internal volume roughly $0.5\,\text{m}^3$ and surface
area $A \sim 3\,\text{m}^2$. The insulation is polyurethane foam, $5\,
\si{\centi\meter}$ thick, with thermal conductivity $k \approx
0.025\,\text{W}/(\text{m}\cdot\text{K})$. The temperature drop across
the wall is $18\,\si{\kelvin}$. The conduction heat leak is
$\dot{Q}_{\text{leak}} \approx k A \Delta T / L = 0.025 \times 3
\times 18 / 0.05 = 27\,\text{W}$.

Door openings and warm items add intermittently; estimate an average
contribution comparable to the steady leak, so $\dot{Q}_{\text{cold}}
\approx 50\,\text{W}$.

A household refrigerator running on a vapour-compression cycle with
condenser at $\sim 35\,\degC = 308\,\si{\kelvin}$ and evaporator at
$\sim -5\,\degC = 268\,\si{\kelvin}$ has an ideal Carnot coefficient of
performance $\text{COP}_{\text{Carnot}} = T_c/(T_h - T_c) = 268/40
\approx 6.7$. A real refrigerator achieves perhaps $40\,\%$ of the
Carnot bound, so $\text{COP}_{\text{actual}} \approx 2.5$ to $3$. The
required electrical power is $\dot{W} = \dot{Q}_{\text{cold}}/\text{COP}
\approx 50/2.7 \approx 18\,\text{W}$.

Over a $24\,\text{h}$ day, the energy consumption is $18 \times 24
\approx 0.43\,\text{kWh}/\text{day}$, or roughly $160\,\text{kWh}/
\text{year}$.

Manufacturers' specifications for ENERGY STAR refrigerators of this
size cluster between $300$ and $450\,\text{kWh}/\text{year}$
\cite{gen:mackay2009}. Our estimate is a factor of two low; the
discrepancy is dominated by larger units, taller temperature
differences in real cycles, defrost cycles, and warmer ambient.

The postmortem: the estimate was honest about the conduction term and
fragile on the door-opening contribution and the cycle efficiency. The
working calculation gets the order of magnitude (tens of watts of
electrical demand) right, which is what the next-stage design
discussion needs.
\end{estimation}

The refrigerator pumps heat $Q_c$ out of the cold cabinet and the
compressor adds work $W$; the condenser then rejects the sum
$Q_h = Q_c + W$ to the kitchen, as
Figure~\ref{fig:vol04ch01:refrigerator-sankey} draws to scale for the
estimate above. The script \texttt{code/refrigerator\_cop.py} reads a
week of logged consumption and cabinet temperatures from
\texttt{data/refrigerator\_power.csv} and returns the measured
coefficient of performance against the Carnot bound; it is the
analysis engine for the chapter project. Property data for the working
gases of this chapter is collected in \texttt{data/gas\_properties.csv}.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-refrigerator-sankey.tex}

\subsection{Polytropic compression with heat loss}

The isothermal and adiabatic compressions of the first worked example
are the two idealised limits. A real compressor sits between them: the
gas heats as it is squeezed, and some of that heat leaks through the
cylinder walls before the stroke finishes. The path is captured by the
\engterm{polytropic} relation $pV^n = \text{const}$, where the exponent
$n$ slides between the isothermal value $n = 1$ and the adiabatic value
$n = \gamma$, and can exceed $\gamma$ when the gas loses heat faster
than an adiabat would allow. Figure~\ref{fig:vol04ch01:polytropic} draws
the family of polytropic paths leaving one start state.

Take one mole of air ($\gamma = 1.4$, $R = 8.314\,\text{J}/(\text{mol}
\cdot\si{\kelvin})$, $c_v = 5R/2$) compressed from $p_1 = 100\,\text{kPa}$,
$T_1 = 300\,\si{\kelvin}$ to $p_2 = 600\,\text{kPa}$ along a polytropic
path with $n = 1.25$, a value typical of a single-stage reciprocating
compressor with modest wall cooling. The end temperature follows from
the polytropic relation written in pressure and temperature,
$T_2 = T_1 (p_2/p_1)^{(n-1)/n} = 300 \times 6^{0.25/1.25} =
300 \times 6^{0.2} = 300 \times 1.431 = 429\,\si{\kelvin}$. The boundary
work on the gas for a polytropic process with $n \neq 1$ is
\[
W = \frac{nR(T_2 - T_1)}{n - 1}
  = \frac{1 \times 8.314 \times (429 - 300)}{0.25}
  = +4290\,\text{J},
\]
positive because work is done on the gas. The change in internal energy
is $\Delta U = c_v(T_2 - T_1) = 2.5 \times 8.314 \times 129 =
+2681\,\text{J}$. The heat is then fixed by the first law,
$Q = \Delta U - W = 2681 - 4290 = -1609\,\text{J}$: the gas rejects
$1609\,\text{J}$ to the cylinder walls. The negative heat is the
signature of the polytropic middle ground; an adiabat would reject none,
an isotherm would reject all of the $4290\,\text{J}$ of work as heat.
The entropy change of the gas is
$\Delta S = c_p \ln(T_2/T_1) - R \ln(p_2/p_1) =
3.5 \times 8.314 \times \ln(1.431) - 8.314 \times \ln 6 =
10.4 - 14.9 = -4.5\,\text{J}/\si{\kelvin}$, negative because the gas is
cooled and compressed. The lesson the polytropic exponent carries is
that a designer who reports compressor work must state the path: the
same end pressures span a range of work from the isothermal floor to the
adiabatic ceiling, and the real machine is found by measuring $n$.

\subsection{Transient filling of an evacuated tank}

The steady-state open-system balance covers process equipment that
runs continuously. A tank being charged is a transient open system, and
the unsteady form of the first law gives a result that surprises the
first-time reader. An insulated, rigid, evacuated tank of volume $V$ is
connected through a valve to a large supply line carrying air at
constant pressure $p_L$ and temperature $T_L$. The valve is opened, air
rushes in, and the valve is shut when the tank pressure reaches the line
pressure. We want the final temperature of the air in the tank.

The unsteady first law for the tank as control volume, with no heat
transfer (insulated), no shaft work, and no work other than the flow
work already inside enthalpy, is
\[
\frac{dU_{cv}}{dt} = \dot{m}_{\text{in}}\, h_L,
\]
where $h_L$ is the specific enthalpy of the air in the line, constant
throughout the fill. Integrate from the empty initial state ($U_{cv} =
0$, $m_{cv} = 0$) to the final state ($m_2$, $u_2$):
$m_2 u_2 = m_2 h_L$, so $u_2 = h_L$. The internal energy per unit mass
of the trapped gas equals the enthalpy per unit mass of the supply. For
an ideal gas with constant specific heats, $u_2 = c_v T_2$ and $h_L =
c_p T_L$, so
\[
T_2 = \frac{c_p}{c_v} T_L = \gamma\, T_L.
\]
Filling an evacuated tank from a line at $T_L = 300\,\si{\kelvin}$ with
air ($\gamma = 1.4$) leaves the trapped gas at $T_2 = 420\,\si{\kelvin}$,
a $120\,\si{\kelvin}$ rise with no heat added and no shaft work done. The
flow work $p_L v_L$ that pushed each parcel across the boundary is
converted to internal energy once the parcel is inside and brought to
rest; that is the entire mechanism. Compressed-gas cylinders warm
audibly when filled for the same reason, and the warming must be allowed
to dissipate before the true charge mass is read, or the cylinder reads
short once it cools. The result is independent of the tank volume and of
the line pressure; it depends only on $\gamma$ and the line temperature.

\subsection{A nozzle: enthalpy traded for speed}

Every worked example so far has discarded the kinetic-energy term in the
open-system balance, justified for turbines, pumps, and heat exchangers
where the stream barely changes speed. A nozzle is the equipment that
breaks that habit. Figure~\ref{fig:vol04ch01:nozzle} draws a control
volume around a nozzle carrying air from a high-pressure plenum to a low
exit pressure. With no shaft work, no heat transfer over the short
residence time, and the potential term negligible, the steady-flow first
law collapses to a balance between enthalpy and kinetic energy,
\[
h_1 + \tfrac12 V_1^2 = h_2 + \tfrac12 V_2^2.
\]
Air enters at $T_1 = 500\,\si{\kelvin}$ with a low approach velocity
$V_1 = 30\,\text{m}/\text{s}$ and expands to $T_2 = 380\,\si{\kelvin}$.
Treating air as ideal with $c_p = 1.005\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$, the enthalpy drop is $h_1 - h_2 = c_p (T_1 - T_2) =
1.005 \times 120 = 120.6\,\text{kJ}/\text{kg}$. Solving the balance for
the exit speed,
\[
V_2 = \sqrt{V_1^2 + 2(h_1 - h_2)}
    = \sqrt{30^2 + 2 \times 120{,}600}
    = \sqrt{900 + 241{,}200}
    = 492\,\text{m}/\text{s}.
\]
The inlet kinetic term ($900\,\text{m}^2/\text{s}^2$) is a third of a
percent of the enthalpy drop ($241{,}200\,\text{m}^2/\text{s}^2$), which
is why the approach velocity is usually dropped and the exit speed read
straight off the enthalpy drop as $V_2 \approx \sqrt{2 c_p (T_1 - T_2)}$.
The exit speed sits just under the local speed of sound for air at
$380\,\si{\kelvin}$ (about $391\,\text{m}/\text{s}$), so the flow has
reached the sonic limit of a simple converging nozzle; pushing it
supersonic needs the converging-diverging geometry that Volume V's
compressible-flow chapter develops. Running the same balance backward
describes a diffuser, which trades the speed of an oncoming stream for a
pressure rise, the job of the inlet of a jet engine before the air ever
reaches the compressor.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-nozzle.tex}

\subsection{Net work of an air-standard Brayton cycle}

The gas turbine of the second case study runs on the Brayton cycle:
isentropic compression, constant-pressure heat addition, isentropic
expansion, constant-pressure rejection.
Figure~\ref{fig:vol04ch01:brayton-ts} draws the cycle on
temperature-entropy axes. The cold-air-standard analysis (constant
specific heats, air throughout) gives the net work in closed form and
exposes the role of the pressure ratio. Take air ($\gamma = 1.4$,
$c_p = 1.005\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$) entering the
compressor at $T_1 = 290\,\si{\kelvin}$, a pressure ratio
$r_p = p_2/p_1 = 12$, and a turbine inlet temperature $T_3 =
1450\,\si{\kelvin}$ set by the metallurgy of the first-stage blades.

The isentropic temperature ratio across either pressure-matched leg is
$T_2/T_1 = T_3/T_4 = r_p^{(\gamma-1)/\gamma} = 12^{0.2857} = 2.034$. The
compressor exit and turbine exit temperatures follow:
\[
T_2 = 290 \times 2.034 = 590\,\si{\kelvin}, \qquad
T_4 = \frac{1450}{2.034} = 713\,\si{\kelvin}.
\]
The compressor work absorbed is $w_c = c_p (T_2 - T_1) = 1.005 \times
300 = 301\,\text{kJ}/\text{kg}$; the turbine work delivered is $w_t =
c_p (T_3 - T_4) = 1.005 \times 737 = 741\,\text{kJ}/\text{kg}$. The net
work is the difference,
\[
w_{\text{net}} = w_t - w_c = 741 - 301 = 440\,\text{kJ}/\text{kg},
\]
and the ratio of compressor work to turbine work, the
\engterm{back-work ratio} $w_c/w_t = 0.41$, is the warning that a gas
turbine spends a large fraction of its turbine output running its own
compressor. A steam plant's pump took less than two percent of the
turbine work; a gas turbine's compressor takes forty. The heat added in
the combustor is $q_{\text{in}} = c_p (T_3 - T_2) = 1.005 \times 860 =
864\,\text{kJ}/\text{kg}$, so the thermal efficiency is
\[
\eta = \frac{w_{\text{net}}}{q_{\text{in}}} = \frac{440}{864} = 0.51,
\]
which the algebra also delivers as $\eta = 1 - r_p^{-(\gamma-1)/\gamma} =
1 - 1/2.034 = 0.51$, a function of the pressure ratio alone for the
cold-air-standard cycle. The turbine inlet temperature does not appear
in the efficiency, but it sets the net work per unit mass, and a high
$T_3$ is what keeps the engine small for its power. The back-work ratio
is the reason a gas turbine fails to start, or stalls, if the turbine
section underperforms even slightly: most of its gross output is already
committed to the compressor.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-brayton-ts.tex}

\subsection{The isentropic efficiency of a real turbine}

The Carnot and Brayton bounds describe reversible machines. A real
turbine generates entropy through blade-passage friction and turbulent
mixing, so its actual exit state lies to the right of the isentropic exit
on the temperature-entropy plane, at a higher temperature and a smaller
enthalpy drop. The \engterm{isentropic efficiency} measures the gap:
\[
\eta_s = \frac{w_{\text{actual}}}{w_{\text{ideal}}}
       = \frac{h_3 - h_{4,\text{actual}}}{h_3 - h_{4s}},
\]
the ratio of the work the real turbine delivers to the work the
reversible turbine would deliver between the same inlet state and the
same exit pressure. Take the turbine leg of the Brayton example above,
with inlet $T_3 = 1450\,\si{\kelvin}$ and an isentropic exit
$T_{4s} = 713\,\si{\kelvin}$, hence an ideal work of $w_{\text{ideal}} =
c_p (T_3 - T_{4s}) = 1.005 \times 737 = 741\,\text{kJ}/\text{kg}$. A
well-built turbine has $\eta_s = 0.88$, so the actual work is
$w_{\text{actual}} = 0.88 \times 741 = 652\,\text{kJ}/\text{kg}$, and the
real exit temperature is higher than the ideal,
\[
T_{4,\text{actual}} = T_3 - \frac{w_{\text{actual}}}{c_p}
  = 1450 - \frac{652}{1.005} = 801\,\si{\kelvin},
\]
some $88\,\si{\kelvin}$ above the isentropic exit. The entropy generated
per unit mass in the expansion is the rise in the gas entropy between the
isentropic and actual exits at the same pressure, $s_{\text{gen}} =
c_p \ln(T_{4,\text{actual}}/T_{4s}) = 1.005 \times \ln(801/713) =
0.117\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$. The shortfall in work,
$741 - 652 = 89\,\text{kJ}/\text{kg}$, is close to the Gouy-Stodola
estimate $T_0 s_{\text{gen}} = 290 \times 0.117 = 34\,\text{kJ}/\text{kg}$
only if the lost work is reckoned against the ambient dead state; the
larger raw shortfall is measured against the turbine exit, not against
ambient, and the difference between the two figures is the exergy still
carried by the hotter-than-ideal exhaust, recoverable downstream. The
isentropic efficiency is the single number a turbine vendor quotes and
the single number that, fed back through this calculation, tells the
plant engineer how much of the cycle's promise the hardware actually
keeps.

\subsection{The polytropic work and heat relations}

The polytropic worked example earlier carried the work formula without
deriving it. The derivation is short and exposes why the same end states
span a whole range of work and heat depending on the path. A polytropic
process is any quasi-static process for which $pV^n = C$ holds with $n$
constant. The boundary work on the gas is the integral $W = -\int p\,
\mathrm{d}V$. Substituting $p = C V^{-n}$,
\[
W = -\int_{V_1}^{V_2} C V^{-n}\,\mathrm{d}V
  = -C\,\frac{V^{1-n}}{1-n}\bigg|_{V_1}^{V_2}
  = \frac{C V_2^{1-n} - C V_1^{1-n}}{n-1}
\]
for $n \neq 1$. Because $C = p_1 V_1^n = p_2 V_2^n$, the products
$C V_2^{1-n} = p_2 V_2$ and $C V_1^{1-n} = p_1 V_1$, so the work
collapses to
\[
W = \frac{p_2 V_2 - p_1 V_1}{n-1} = \frac{nR(T_2 - T_1)}{n-1}
\]
per mole, using $pV = nRT$ at each end. The adiabatic result is the
special case $n = \gamma$, and the case $n = 1$ is excluded because the
integral of $V^{-1}$ is a logarithm: at $n = 1$ the process is
isothermal and $W = -nRT\ln(V_2/V_1)$ instead. The heat follows from the
first law with $\Delta U = n c_v (T_2 - T_1)$,
\[
Q = \Delta U - W = n c_v(T_2 - T_1) - \frac{nR(T_2 - T_1)}{n-1}
  = n c_v\,(T_2 - T_1)\,\frac{\gamma - n}{n-1},
\]
where the last step uses $R = c_p - c_v = (\gamma - 1)c_v$. The sign of
the heat is read straight off the factor $(\gamma - n)/(n-1)$: for a
compression with $1 < n < \gamma$ the factor is positive while
$T_2 > T_1$, so $Q > 0$ when written this way for heat added, which for
a cooled compressor means the gas rejects heat (the bracket carries the
sign of $T_2 - T_1$ against the convention). The reader who keeps the
algebra rather than the verbal rule recovers the earlier numerical
example, $n = 1.25$ and $\gamma = 1.4$ giving a small negative heat, and
sees that an isotherm ($n = 1$) dumps all the work as heat while an
adiabat ($n = \gamma$) dumps none. The polytropic exponent is the single
parameter that slides the process between those limits, and the work and
heat are explicit functions of it.

\subsection{An isobaric heat addition with boundary work}

A frequent first-law confusion is to set the heat equal to the
internal-energy change when the volume is free to move. Take
$0.20\,\text{kg}$ of air ($c_p = 1.005$, $c_v = 0.718\,\text{kJ}/
(\text{kg}\cdot\si{\kelvin})$, $R = 287\,\text{J}/(\text{kg}\cdot
\si{\kelvin})$) in a piston-cylinder under a constant load that holds the
pressure at $300\,\text{kPa}$. Heat is added until the temperature rises
from $T_1 = 300\,\si{\kelvin}$ to $T_2 = 500\,\si{\kelvin}$. The heat
added at constant pressure is the enthalpy rise,
\[
Q = m c_p (T_2 - T_1) = 0.20 \times 1.005 \times 200 = 40.2\,\text{kJ}.
\]
The gas expands, doing work on the piston. The boundary work on the gas
is $W = -p\,\Delta V = -mR(T_2 - T_1) = -0.20 \times 287 \times 200 =
-11.5\,\text{kJ}$, negative because the gas pushes the piston out. The
internal-energy change checks the first law:
$\Delta U = Q + W = 40.2 - 11.5 = 28.7\,\text{kJ}$, which equals
$m c_v (T_2 - T_1) = 0.20 \times 0.718 \times 200 = 28.7\,\text{kJ}$
independently, as it must for an ideal gas whose internal energy depends
on temperature alone. Of the $40.2\,\text{kJ}$ of heat, $28.7\,\text{kJ}$
raised the internal energy and $11.5\,\text{kJ}$ left as work against the
piston; the split is exactly $c_v/c_p$ to internal energy and $R/c_p$ to
work, the ratio $0.714$ to $0.286$ that every constant-pressure heating
of a diatomic gas obeys.

\subsection{Throttling a real gas and the Joule-Thomson inversion}

A gas forced through a valve or a porous plug drops in pressure with no
shaft work and, over the short passage, no appreciable heat transfer.
The steady-flow balance with $\dot{W}_s = 0$, $\dot{Q} = 0$, and
negligible velocity change reduces to $h_1 = h_2$: throttling is an
\engterm{isenthalpic} process. For an ideal gas, enthalpy depends only on
temperature, so $h_1 = h_2$ forces $T_1 = T_2$ and a throttle does
nothing to the temperature. A real gas is the interesting case. The
temperature change with pressure at constant enthalpy is the
\engterm{Joule-Thomson coefficient} $\mu_{JT} = (\partial T/\partial
p)_h$, and a thermodynamic identity gives it as
\[
\mu_{JT} = \frac{1}{c_p}\!\left[T\!\left(\pdv{v}{T}\right)_p - v\right],
\]
which is positive when the gas cools on throttling and negative when it
warms. Take carbon dioxide near room conditions, where $(\partial v/
\partial T)_p$ exceeds $v/T$, so $\mu_{JT} > 0$ and the gas cools: a
pressure drop of $5\,\text{MPa}$ with a coefficient of roughly
$1.1\,\si{\kelvin}/\text{bar}$ near ambient cools the stream by tens of
kelvin, the mechanism behind the carbon-dioxide fire extinguisher's
frosted horn and behind the Linde process that liquefies air. Above the
\engterm{inversion temperature} the coefficient changes sign and the same
throttle warms the gas; hydrogen and helium sit below their inversion
temperatures only when pre-cooled, which is why a hydrogen cylinder
vented to atmosphere warms rather than cools. The lesson is that the
ideal-gas result ($T$ unchanged) is the boundary case between cooling and
warming, and which side a real gas falls on is set by the sign of
$T(\partial v/\partial T)_p - v$. The locus where the sign flips is the
\engterm{inversion curve} of Figure~\ref{fig:vol04ch01:jt-inversion}, a closed
loop on the pressure-temperature plane that separates the cooling interior from
the warming exterior and meets the temperature axis at the maximum inversion
temperature. Volume's chapter on equations of state evaluates the coefficient
from a real-gas model; the present point is that the isenthalpic condition
$h_1 = h_2$ is the whole of the first-law input.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-jt-inversion.tex}

A number closes the throttling case and shows the mechanism in joules. A carbon
dioxide stream is throttled from $p_1 = 6.0\,\text{MPa}$ to $p_2 =
1.0\,\text{MPa}$ near room temperature, where its Joule-Thomson coefficient
averages about $\mu_{JT} \approx 1.1\,\si{\kelvin}/\text{bar}$ over the range. The
pressure drop is $\Delta p = -50\,\text{bar}$, so the temperature change is
$\Delta T \approx \mu_{JT}\,\Delta p = 1.1 \times (-50) = -55\,\si{\kelvin}$: the
stream leaves some fifty-five kelvin colder than it entered, with no shaft work and
no heat added. The cooling is bought from the internal energy the gas spends
pulling its own molecules apart against their attraction as they spread out, the
positive $(\partial U/\partial V)_T$ the earlier thermodynamic equation of state
located for a real gas. An ideal gas, with $(\partial U/\partial V)_T = 0$, has
nothing to spend and stays at its inlet temperature, which is the flat centre-line
the inversion curve is drawn around. The carbon-dioxide fire extinguisher exploits
exactly this drop: the liquid flashes to cold gas at the horn and the throttle
chills what remains toward the sublimation point, coating the horn in frost.

\subsection{Charging a tank that is not empty}

The transient-filling example earlier started from an evacuated tank and
found the trapped gas at $\gamma T_L$. The more common field situation is
topping up a tank that already holds gas, and the unsteady first law
handles it with one more term. An insulated rigid tank of volume
$V = 0.5\,\text{m}^3$ holds air at $p_1 = 200\,\text{kPa}$,
$T_1 = 300\,\si{\kelvin}$; it is connected to a line at
$p_L = 1.0\,\text{MPa}$, $T_L = 300\,\si{\kelvin}$ and charged until the
tank reaches $p_2 = 1.0\,\text{MPa}$. The unsteady energy balance for the
tank, insulated and with no shaft work, is
\[
m_2 u_2 - m_1 u_1 = (m_2 - m_1)\,h_L,
\]
the change in stored internal energy equalling the enthalpy carried in by
the added mass. With $u = c_v T$, $h_L = c_p T_L$, and the masses from the
ideal-gas law $m = pV/(RT)$,
\[
m_1 = \frac{200{,}000 \times 0.5}{287 \times 300} = 1.16\,\text{kg},
\qquad
m_2 = \frac{1{,}000{,}000 \times 0.5}{287 \times T_2},
\]
where $T_2$ is unknown and appears in $m_2$. Substituting and solving the
balance $m_2 c_v T_2 - m_1 c_v T_1 = (m_2 - m_1) c_p T_L$ for $T_2$ gives,
after cancelling $c_v$ and writing $\gamma = c_p/c_v$,
\[
T_2 = \frac{p_2 V/R}{\,m_1 (T_1 - \gamma T_L)/T_1 + p_2 V \gamma T_L/(R\,
T_1 T_L)\,}
\]
which is cleaner solved numerically: $m_1 c_v T_1 = 1.16 \times 718
\times 300 = 250\,\text{kJ}$, and the balance reads
$m_2 (718\,T_2) - 250{,}000 = (m_2 - 1.16)(1005 \times 300)$ with
$m_2 = 1{,}742/T_2$ (in kg when $T_2$ is in kelvin, from $m_2 = p_2 V/
(R T_2)$). Solving, $T_2 \approx 380\,\si{\kelvin}$ and
$m_2 \approx 4.58\,\text{kg}$. The tank warms by $80\,\si{\kelvin}$,
less than the $120\,\si{\kelvin}$ of the evacuated case because the
resident gas, already at the tank temperature, dilutes the enthalpy
brought in by the charge. The same warming must dissipate before the
true stored mass is read off the pressure gauge, the practical warning
the evacuated case raised, now quantified for a partly full tank.

\subsection{Free expansion of an ideal gas into a vacuum}

The worked examples so far have moved a piston quasi-statically. A free
expansion does the opposite: a gas held in one half of an insulated rigid
vessel is released through a valve into an evacuated other half, and it
rushes in with no restraining piston to push against. The first law for the
whole rigid insulated vessel is stark. No heat crosses the insulated wall, so
$Q = 0$; no boundary moves, so the gas does no work on anything outside the
vessel, $W = 0$; hence $\Delta U = 0$. For an ideal gas, whose internal
energy depends on temperature alone, $\Delta U = 0$ forces $\Delta T = 0$:
the gas ends at the same temperature it began, with no work extracted and no
heat exchanged. Take one mole of air initially at $300\,\si{\kelvin}$ in
$V_1 = 10\,\si{\liter}$ released into a total volume $V_2 = 20\,\si{\liter}$.
The final temperature is $300\,\si{\kelvin}$, unchanged.

The entropy change is not zero, and computing it shows why the state-function
machinery is worth its keep. The actual process is violent and far from
equilibrium, so $\delta Q_{\text{rev}}/T$ cannot be integrated along the real
path. Entropy is a state function, so any reversible path between the same end
states serves. A reversible isothermal expansion from $V_1$ to $V_2$ at
$300\,\si{\kelvin}$ connects them, and along it
\[
\Delta S = nR\ln\frac{V_2}{V_1} = 1 \times 8.314 \times \ln 2 =
+5.76\,\text{J}/\si{\kelvin}.
\]
The gas entropy rose by $5.76\,\text{J}/\si{\kelvin}$ while no heat entered
the isolated vessel, so the entropy generation equals the whole of it:
$\dot S_{\text{gen}} = 5.76\,\text{J}/\si{\kelvin}$ for the event. The free
expansion is the cleanest illustration of an irreversible process that
exchanges neither heat nor work yet generates entropy, and of the rule that
entropy is computed along a reversible substitute path while heat and work
are not.

\subsection{A reversible-adiabatic and an irreversible compression compared}

Two compressors take the same air from the same inlet to the same final
pressure, one reversible and adiabatic, the other adiabatic with
irreversibility, and the comparison isolates the cost of entropy generation
in joules. One mole of air ($\gamma = 1.4$, $c_v = 5R/2$, $c_p = 7R/2$) enters
at $p_1 = 100\,\text{kPa}$, $T_1 = 300\,\si{\kelvin}$ and is compressed to
$p_2 = 400\,\text{kPa}$. The reversible adiabatic machine follows
$T_{2s} = T_1 (p_2/p_1)^{(\gamma-1)/\gamma} = 300 \times 4^{0.2857} =
446\,\si{\kelvin}$ and absorbs $w_s = c_v(T_{2s} - T_1) = 2.5 \times 8.314
\times 146 = 3030\,\text{J}$ with $\Delta S = 0$.

The real machine has an isentropic efficiency $\eta_s = 0.80$, so it absorbs
more work for the same pressure rise, $w = w_s/\eta_s = 3030/0.80 =
3790\,\text{J}$, and the extra work appears as a hotter exit. From the energy
balance $w = c_v(T_2 - T_1)$ the actual exit temperature is $T_2 = 300 +
3790/(2.5 \times 8.314) = 482\,\si{\kelvin}$, some $36\,\si{\kelvin}$ above
the reversible exit at the same pressure. The entropy generated is the gas
entropy rise between the two exits at $p_2$,
\[
s_{\text{gen}} = c_p\ln\frac{T_2}{T_{2s}} = 3.5 \times 8.314 \times
\ln\frac{482}{446} = +2.26\,\text{J}/\si{\kelvin}.
\]
Charging that generation at the Gouy-Stodola rate against a
$T_0 = 300\,\si{\kelvin}$ environment, the lost work is $T_0 s_{\text{gen}} =
300 \times 2.26 = 678\,\text{J}$, which matches the raw work shortfall
$3790 - 3030 = 760\,\text{J}$ to the accuracy of the cold-air-standard
rounding, the residue being the exergy still carried by the hotter exit gas.
The irreversible machine spent an extra $760\,\text{J}$ to reach the same
pressure, and the second law accounts for nearly all of it as
$T_0 s_{\text{gen}}$.

\subsection{Entropy generated heating a house with a resistance heater
versus a heat pump}

A first-law-only accounting rates a resistance heater as perfect: every joule
of electricity becomes a joule of heat in the room. The second law rates it
as wasteful, and the entropy balance shows the size of the waste. A room held
at $T_h = 294\,\si{\kelvin}$ (21\,$^{\circ}\mathrm{C}$) loses
$\dot Q_h = 3\,\text{kW}$ to the outdoors at $T_c = 273\,\si{\kelvin}$
(0\,$^{\circ}\mathrm{C}$), and that heat must be supplied. A resistance heater
draws $\dot W = 3\,\text{kW}$ of electricity, a high-grade form of work with
zero entropy, and dumps it as $3\,\text{kW}$ of heat into the room at
$T_h$. The entropy generation rate of the conversion is the entropy delivered
to the room less the entropy carried in by the work (none),
\[
\dot S_{\text{gen}} = \frac{\dot Q_h}{T_h} - 0 =
\frac{3000}{294} = 10.2\,\text{W}/\si{\kelvin},
\]
and the lost work is $T_c \dot S_{\text{gen}} = 273 \times 10.2 =
2790\,\text{W}$, nearly the whole electrical input destroyed as work
potential. A heat pump delivering the same $3\,\text{kW}$ to the room while
drawing heat from the outdoor air at $T_c$ needs, at the Carnot limit, only
$\dot W = \dot Q_h (1 - T_c/T_h) = 3000 \times (1 - 273/294) = 214\,\text{W}$,
a factor of fourteen less electricity, because it moves low-grade ambient heat
rather than degrading high-grade work into it. The two devices satisfy the
first law identically; the second law separates them, and the entropy balance
puts a wattage on the difference. The resistance heater is the most
thermodynamically wasteful way to use electricity ever sold, and it remains in
service only because its hardware is a coil of wire while the heat pump's is a
sealed refrigeration cycle. The comparison is really about the grade of energy,
which Figure~\ref{fig:vol04ch01:energy-grade} draws: electrical work is nearly
pure exergy, heat delivered near ambient is nearly pure anergy, and the resistance
heater converts the first into the second, conserving the joules while destroying
the work potential.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-energy-grade.tex}

\subsection{Adiabatic flame temperature as a first-law balance}

Combustion is a first-law accounting problem once the chemical energy is
booked as an enthalpy of reaction. A burner mixes methane with the
stoichiometric quantity of air and burns it with no heat loss to the
surroundings; the products carry away the entire enthalpy released, which
fixes their temperature, the \engterm{adiabatic flame temperature}. The
balance is the constant-pressure statement that the enthalpy of the cold
reactants equals the enthalpy of the hot products, because the burner is
adiabatic and does no shaft work,
\[
H_{\text{reactants}}(T_0) = H_{\text{products}}(T_{\text{ad}}).
\]
Splitting the product enthalpy into the value at the reference temperature
plus the sensible rise gives the working form: the heat of combustion warms
the products from $T_0$ to $T_{\text{ad}}$. Methane releases a lower heating
value of about $802\,\text{kJ}$ per mole burned. Burning one mole of methane
with the stoichiometric $2$ moles of oxygen and the $7.52$ moles of nitrogen
that accompany it in air produces $1$ mole of carbon dioxide, $2$ moles of
water vapour, and the $7.52$ moles of inert nitrogen carried through, a total
of $10.52$ moles of hot product. Taking a representative mean molar heat
capacity of the product mixture as $\bar c_p \approx 40\,\text{J}/(\text{mol}
\cdot\si{\kelvin})$ over the temperature range, the sensible balance is
\[
802{,}000\,\text{J} = (10.52\,\text{mol})(40\,\text{J}/(\text{mol}\cdot
\si{\kelvin}))\,(T_{\text{ad}} - 298\,\si{\kelvin}),
\]
which solves to $T_{\text{ad}} - 298 = 802{,}000/421 = 1906\,\si{\kelvin}$,
so $T_{\text{ad}} \approx 2200\,\si{\kelvin}$. The measured adiabatic flame
temperature of stoichiometric methane in air is near $2230\,\si{\kelvin}$, so
the one-line balance with a constant mean heat capacity lands within two
percent. The nitrogen that took no part in the reaction carries more than
seventy percent of the product moles and so does most of the work of limiting
the flame temperature; running the same fuel in pure oxygen removes that
ballast and pushes the flame past $3000\,\si{\kelvin}$, the reason an
oxy-fuel torch cuts steel that an air-fuel flame only warms. The first law
sized the flame with no kinetics and no equilibrium chemistry, the same way
it sized the steam plant's streams.

\subsection{Optimal intercooling: the pressure ratio that splits the work}

The compressed-air plant compressed in three stages of equal pressure ratio and
cooled between them, and the choice of equal ratios was asserted rather than
justified. The first law justifies it. Take a two-stage compression of air from
$p_1$ to $p_3$ with an intercooler that returns the gas to the inlet temperature
$T_1$ between the stages, each stage isentropic with the same efficiency, and ask
what intermediate pressure $p_2$ minimises the total work. The per-stage
isentropic work on the gas is $c_p T_1[(p_{\text{out}}/p_{\text{in}})^{k} - 1]$
with $k = (\gamma-1)/\gamma$, and because the intercooler restores $T_1$ before
the second stage, both stages start from the same temperature. The total is
\[
w = c_p T_1\!\left[\left(\frac{p_2}{p_1}\right)^{k} +
\left(\frac{p_3}{p_2}\right)^{k} - 2\right].
\]
Differentiate with respect to $p_2$ and set the derivative to zero. The two
terms pull in opposite directions, and the stationary point falls where
$(p_2/p_1)^{k}/p_2 = (p_3/p_2)^{k}/p_2$ after the algebra collapses, which
reduces to $p_2 = \sqrt{p_1 p_3}$: the intermediate pressure is the geometric
mean of the end pressures, so each stage carries the same pressure ratio
$\sqrt{p_3/p_1}$. For the three-stage train of the storage plant the same
argument, extended to $N$ stages, gives equal ratios $r = (p_3/p_1)^{1/N}$ per
stage, which is why the plant used $60^{1/3} = 3.91$ throughout. Put numbers to
the two-stage case: air from $p_1 = 100\,\text{kPa}$ to $p_3 = 900\,\text{kPa}$,
$T_1 = 300\,\si{\kelvin}$, $c_p = 1.005\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$.
Single-stage isentropic work is $c_p T_1(9^{0.2857} - 1) = 1.005 \times 300
\times 0.874 = 264\,\text{kJ}/\text{kg}$. Two stages at the optimum ratio
$\sqrt{9} = 3$ each cost $2 c_p T_1(3^{0.2857} - 1) = 2 \times 1.005 \times 300
\times 0.369 = 223\,\text{kJ}/\text{kg}$, a saving of sixteen percent, all of it
the heat the intercooler carried away so the second stage compressed cooler,
denser gas. Each further stage claws back less; the intercooler count is set by
where the saving no longer pays for the extra heat exchanger, the classic balance
the plant designer strikes.

\subsection{Entropy generated in a counterflow heat exchanger}

A heat exchanger transfers heat between two streams without mixing them, and the
first law alone rates a well-insulated exchanger as loss-free: every joule
leaving the hot stream enters the cold one. The second law finds the loss the
first law misses, the entropy generated because the transfer crosses a finite
temperature difference. Take a counterflow exchanger in which a hot air stream of
$\dot m_h = 1.0\,\text{kg}/\text{s}$ cools from $T_{h,\text{in}} =
600\,\si{\kelvin}$ to $T_{h,\text{out}} = 400\,\si{\kelvin}$ while heating a cold
air stream of $\dot m_c = 1.0\,\text{kg}/\text{s}$ from $T_{c,\text{in}} =
300\,\si{\kelvin}$, with $c_p = 1.005\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$ for
both. The heat given up by the hot stream is $\dot Q = \dot m_h c_p(T_{h,\text{in}}
- T_{h,\text{out}}) = 1.0 \times 1.005 \times 200 = 201\,\text{kW}$, and the same
heat raises the cold stream by $\Delta T_c = \dot Q/(\dot m_c c_p) =
201/1.005 = 200\,\si{\kelvin}$, so $T_{c,\text{out}} = 500\,\si{\kelvin}$. No heat
crosses the outer casing, so the entropy generation rate is the net rise in the
two streams' entropies,
\[
\dot S_{\text{gen}} = \dot m_h c_p \ln\frac{T_{h,\text{out}}}{T_{h,\text{in}}}
+ \dot m_c c_p \ln\frac{T_{c,\text{out}}}{T_{c,\text{in}}}
= 1.005\!\left(\ln\frac{400}{600} + \ln\frac{500}{300}\right).
\]
The hot stream sheds $1.005 \times \ln(2/3) = -0.408\,\text{kW}/\si{\kelvin}$ of
entropy and the cold stream gains $1.005 \times \ln(5/3) =
+0.513\,\text{kW}/\si{\kelvin}$, so $\dot S_{\text{gen}} =
+0.105\,\text{kW}/\si{\kelvin}$, positive as the second law demands. Charged at
the Gouy-Stodola rate against a $T_0 = 300\,\si{\kelvin}$ environment, the lost
work is $T_0 \dot S_{\text{gen}} = 300 \times 0.105 = 31.5\,\text{kW}$, roughly a
sixth of the heat transferred, destroyed purely because the two streams never
touched at equal temperature. Narrowing the approach temperature, the gap between
the streams at the pinch, cuts $\dot S_{\text{gen}}$ toward zero but demands more
heat-transfer area, and the exchanger designer trades capital cost against exergy
destroyed exactly along this curve. The result also explains why a counterflow
arrangement beats a parallel-flow one for the same duty: counterflow holds a
smaller, more uniform temperature gap along the whole length, so it generates less
entropy for the same heat moved.

\subsection{Pressure rise of a sealed rigid vessel from the cyclic relation}

The cyclic relation was derived earlier and its payoff stated in words; a number
makes it concrete and connects it to the relief-sizing the section promised. A
rigid sealed steel bottle is filled with water (nearly incompressible) at
$20\,^{\circ}\mathrm{C}$ and left in the sun, where its temperature rises by
$\Delta T = 30\,\si{\kelvin}$. What pressure does it build? Heating a sealed rigid
vessel raises pressure at constant volume, the derivative $(\partial p/\partial
T)_V$, which the cyclic relation delivers from two routine measurements as
\[
\left(\frac{\partial p}{\partial T}\right)_V = \frac{\alpha}{\kappa_T},
\]
with $\alpha = (1/V)(\partial V/\partial T)_p$ the volumetric thermal expansion
coefficient and $\kappa_T = -(1/V)(\partial V/\partial p)_T$ the isothermal
compressibility. For liquid water near room temperature $\alpha \approx
2.1 \times 10^{-4}\,\si{\kelvin}^{-1}$ and $\kappa_T \approx 4.6 \times
10^{-10}\,\text{Pa}^{-1}$, so
\[
\left(\frac{\partial p}{\partial T}\right)_V = \frac{2.1\times 10^{-4}}
{4.6\times 10^{-10}} = 4.6 \times 10^{5}\,\text{Pa}/\si{\kelvin},
\]
about $4.6\,\text{bar}$ per kelvin. A $30\,\si{\kelvin}$ rise then builds roughly
$1.4 \times 10^{7}\,\text{Pa}$, some $140\,\text{bar}$, if the bottle does not
yield or leak first, which is why a sealed liquid-full vessel with no vapour space
and no relief is a hazard out of all proportion to the mild temperature change.
The designer never heats a sealed vessel to read the slope; the cyclic relation
supplies it from $\alpha$ and $\kappa_T$, both listed in any property table, and
the relief valve is sized to open long before the wall stress from this rate
reaches the yield of the steel. The small liquid compressibility is the villain:
a gas-filled vessel would rise only by $(\partial p/\partial T)_V = p/T$, a
fraction of a bar per kelvin, because a gas is a thousand times more compressible
and absorbs the thermal expansion into a pressure rise a thousand times gentler.

\subsection{Two finite blocks reaching a common temperature}

The copper-block-in-a-lake exercise cooled a small body against an infinite
reservoir. The finite-reservoir version, two equal blocks brought into contact,
shows a subtler entropy accounting and sets up the maximum-work idea that follows.
Two identical copper blocks of mass $m = 1\,\text{kg}$ and specific heat $c =
385\,\text{J}/(\text{kg}\cdot\si{\kelvin})$, one at $T_1 = 400\,\si{\kelvin}$ and
one at $T_2 = 300\,\si{\kelvin}$, are placed in contact inside an insulated box and
allowed to equilibrate with no work extracted. Energy conservation with equal heat
capacities fixes the common temperature as the arithmetic mean, $T_f = (T_1 +
T_2)/2 = 350\,\si{\kelvin}$. The entropy generated is the sum over both blocks,
each computed along a reversible heating or cooling path to $T_f$,
\[
S_{\text{gen}} = mc\ln\frac{T_f}{T_1} + mc\ln\frac{T_f}{T_2}
= mc\ln\frac{T_f^2}{T_1 T_2}
= 385\ln\frac{350^2}{400 \times 300}.
\]
The argument is $122{,}500/120{,}000 = 1.0208$, so $S_{\text{gen}} = 385 \times
0.0206 = 7.9\,\text{J}/\si{\kelvin}$, positive because the final temperature is the
arithmetic mean while the entropy floor sits at the geometric mean $\sqrt{T_1 T_2}
= 346.4\,\si{\kelvin}$. That gap between the arithmetic and geometric means is the
whole of the irreversibility: had a reversible engine run between the blocks
instead of a bare contact, it would have driven them to the geometric mean while
delivering work, generating no entropy. The maximum work recoverable is the
energy that the bare contact wasted,
\[
W_{\max} = mc(T_1 + T_2 - 2\sqrt{T_1 T_2})
= 385(700 - 2\times 346.4) = 385 \times 7.2 = 2.8\,\text{kJ},
\]
the difference between the internal energy the blocks start with and the internal
energy they would hold at the common geometric-mean temperature. The bare contact
threw that $2.8\,\text{kJ}$ away as generated entropy; the number equals
$T_0 S_{\text{gen}}$ only approximately here because the reservoir temperature
itself drifts, but the structure is the same one the Gouy-Stodola theorem carries.
Two finite bodies at different temperatures are a battery of available work, and
letting them equalise without an engine between them spends the battery for
nothing.

\subsection{Maximum work from a finite hot body cooling to ambient}

The finite-block result generalises to the most common exergy question an engineer
meets: a hot mass is available and an ambient environment surrounds it, so how much
work can be extracted as the mass cools to ambient? Take $m = 100\,\text{kg}$ of
steel ($c = 500\,\text{J}/(\text{kg}\cdot\si{\kelvin})$) at $T_1 =
800\,\si{\kelvin}$, an ambient at $T_0 = 300\,\si{\kelvin}$. A reversible engine
operating between the cooling mass and the fixed environment extracts the mass's
exergy, which for a body of temperature-independent heat capacity is
\[
W_{\max} = mc\!\left[(T_1 - T_0) - T_0\ln\frac{T_1}{T_0}\right],
\]
the enthalpy drop $mc(T_1 - T_0)$ discounted by the unavailable part
$T_0\,\Delta S$. Numerically $mc = 50{,}000\,\text{J}/\si{\kelvin}$, the enthalpy
drop is $50{,}000 \times 500 = 25\,\text{MJ}$, and the discount is $50{,}000 \times
300 \times \ln(800/300) = 50{,}000 \times 300 \times 0.981 = 14.7\,\text{MJ}$, so
\[
W_{\max} = 25 - 14.7 = 10.3\,\text{MJ}.
\]
Only $10.3\,\text{MJ}$ of the $25\,\text{MJ}$ of heat the mass gives up is
available as work; the remaining $14.7\,\text{MJ}$ must be dumped to the
environment as the price of the second law, even with a perfect engine. The
available fraction, $41\,\%$ here, climbs toward one as $T_1$ rises far above
ambient and falls toward zero as $T_1$ approaches $T_0$, which is why a
foundry billet cooling from red heat is a serious work source while a warm radiator
a few kelvin above the room is nearly worthless. The same integral, with the sign
reversed, gives the minimum work to cool a body below ambient, the bound a
refrigerator's compressor is measured against.

\subsection*{An applied vignette: the sodium-acetate hand warmer}

A reusable hand warmer is a flat plastic pouch of sodium acetate solution
with a small steel disc inside. Flexing the disc starts a wave of
crystallisation that floods the pouch with warmth in seconds and holds it for
the better part of an hour; dropping the spent pouch in boiling water
recharges it. The device is a phase-change energy store, and the first and
second laws account for it with the same bookkeeping that sized the
compressed-air plant, on a scale that fits in a glove.

The working substance is sodium acetate trihydrate, which melts at
$T_m \approx 58\,^{\circ}\mathrm{C}$ and carries a latent heat of fusion of
about $h_{sf} = 260\,\text{kJ}/\text{kg}$. The recharge step boils the pouch
to melt the crystals fully, leaving a liquid melt. As the pouch cools back to
room temperature the melt does not refreeze; it passes its freezing point and
sits as a \engterm{supersaturated} liquid, a metastable state that holds the
crystallisation enthalpy with no temperature signature, the flat line at left
in Figure~\ref{fig:vol04ch01:handwarmer}. The stored energy is real and
latent: the pouch at room temperature carries the heat of fusion locked in
the metastable liquid, paid for by the boiling water during recharge.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-handwarmer.tex}

Flexing the disc nucleates a crystal, and the crystallisation front sweeps the
pouch. The first law for the pouch, treated as adiabatic over the few seconds
of the snap, sets the released latent heat equal to the sensible warming of
the pouch contents,
\[
m\,h_{sf} = m\,\bar c_p\,(T_m - T_0),
\]
which for a pouch of any mass cancels $m$ and predicts the pouch will heat
until it reaches the melt temperature and then plateau there while the rest of
the liquid solidifies at constant temperature, the latent-heat plateau in the
figure. With a solution specific heat of roughly $\bar c_p \approx 3\,\text{kJ}
/(\text{kg}\cdot\si{\kelvin})$ and a room at $21\,^{\circ}\mathrm{C}$, the
sensible warming to the plateau consumes only $\bar c_p (T_m - T_0) = 3 \times
37 = 111\,\text{kJ}/\text{kg}$ of the $260\,\text{kJ}/\text{kg}$ available,
leaving more than half the stored heat to be released slowly at the plateau as
the front finishes and then as the solid pouch cools, the long tail in the
figure. A $100\,\text{g}$ pouch therefore stores about
$0.10 \times 260 = 26\,\text{kJ}$, enough to deliver tens of watts for the
half-hour or so the plateau and tail last, which matches the felt behaviour.

The second law explains the metastability that makes the device possible.
A supersaturated solution sits in a local free-energy minimum separated from
the crystalline state by a nucleation barrier; the solution is not at
equilibrium, but it cannot reach equilibrium without a seed because forming
the first tiny crystal raises the free energy through surface energy before
the bulk term pays it back. The steel disc, flexed, throws off a few
microscopic crystals from a thin adhered layer and supplies the seed, after
which the bulk free-energy drop drives the front to completion. The recharge,
boiling, redissolves every crystal and erases the seeds, returning the pouch
to the metastable branch. The whole cycle is a store-and-release of latent
heat with the second law guarding the gate: the energy waits, ready and
spontaneous-once-triggered, held back only by the nucleation barrier that the
disc defeats on demand.

\subsection*{An applied vignette: the Peltier drink cooler}

A camping refrigerator that runs off a car's electrical outlet holds no
compressor, no valve, and no circulating fluid; it cools with a thermoelectric
module the size of a coaster, a flat ceramic sandwich of many small
semiconductor pillars wired electrically in series and thermally in parallel.
Passing a direct current through the module pumps heat from one ceramic face to
the other, the \engterm{Peltier effect}, and the first and second laws size the
device with the same refrigerator bookkeeping used for the vapour-compression
cabinet, on hardware with no moving parts.

The module is a heat pump driven by electrical work. Call the cold face $T_c$,
the hot face $T_h$, and the electrical power drawn $\dot W$. The first law for the
module as a control volume is the refrigerator balance: the heat rejected at the
hot face equals the heat lifted from the cold face plus the electrical work,
$\dot Q_h = \dot Q_c + \dot W$. The catch that makes thermoelectric cooling weak
is the coefficient of performance. A good single-stage module reaches a
$\text{COP} = \dot Q_c/\dot W$ of only about $0.4$ to $0.6$ across a
$20\,\si{\kelvin}$ lift, where a vapour-compression cabinet reaches two to three,
because two loss mechanisms fight the Peltier pumping. The current that pumps the
heat also dissipates ohmic heat in the pillars at a rate $I^2 R$, half of which
flows back to the cold face; and the same pillars that carry the current conduct
heat backward down the temperature gradient they are trying to maintain, a leak
$K(T_h - T_c)$ through the module's own thermal conductance. Both losses grow as
the lift $T_h - T_c$ widens, so a single module stalls at a maximum lift of about
$60\,\si{\kelvin}$ where the conduction and Joule leaks exactly cancel the
pumping and $\dot Q_c$ falls to zero.

Put numbers to a drink cooler that must hold a $15\,\text{L}$ box at
$8\,^{\circ}\mathrm{C}$ in a $28\,^{\circ}\mathrm{C}$ car, a lift of
$20\,\si{\kelvin}$. The cabinet leaks heat like the kitchen refrigerator estimated
earlier; scale that estimate to the smaller box and thinner insulation and the
cold load is roughly $\dot Q_c \approx 20\,\text{W}$. At a module
$\text{COP} \approx 0.5$ the electrical draw is $\dot W = \dot Q_c/\text{COP}
\approx 40\,\text{W}$, and the hot face must reject $\dot Q_h = \dot Q_c + \dot W
\approx 60\,\text{W}$, three times the heat it lifts. That rejection is why every
thermoelectric cooler carries a finned heat sink and a fan on its hot side far
larger than the cold box would suggest: the second law forces the hot face to
shed the lifted heat plus the whole of the electrical work, and if the fin cannot
keep $T_h$ down the lift the module can sustain collapses. The device trades the
compressor's efficiency for silence, robustness, and the ability to reverse from
cooling to heating by flipping the current, the reason it survives in car seats,
scientific instruments, and small coolers despite a coefficient of performance a
vapour-compression cycle would be ashamed of. The first law fixed the three
streams; the second law, through the COP, explained why this elegant solid-state
device stays a niche one.

\begin{mastery}
The Carnot efficiency $1 - T_c/T_h$ is the ceiling for an engine that transfers
heat reversibly, which requires infinitely slow heat exchange across zero
temperature difference and so delivers zero power. A real plant must run at finite
speed, and the useful bound is the efficiency at maximum power output rather than
at zero power. Model an \engterm{endoreversible} engine: a reversible Carnot core
that exchanges heat with its reservoirs only across finite conductances, so the
working fluid sits at an internal hot temperature $T_1 < T_h$ and internal cold
temperature $T_2 > T_c$, and the entropy generation lives entirely in the two
heat-exchanger gaps. Writing the heat rates as $\dot Q_h = \kappa_h(T_h - T_1)$
and $\dot Q_c = \kappa_c(T_2 - T_c)$, imposing the reversible core condition
$\dot Q_h/T_1 = \dot Q_c/T_2$, and maximising the power $\dot W = \dot Q_h -
\dot Q_c$ over the internal temperatures yields the \engterm{Curzon-Ahlborn}
efficiency,
\[
\eta_{CA} = 1 - \sqrt{\frac{T_c}{T_h}},
\]
independent of the conductances. For the steam plant's reservoirs
($T_h = 813\,\si{\kelvin}$, $T_c = 319\,\si{\kelvin}$) the Carnot bound is
$0.61$ while $\eta_{CA} = 1 - \sqrt{319/813} = 0.37$, strikingly close to the
$0.40$ the Rankine cycle actually reached. The agreement is partly luck, since
real plants are not endoreversible in detail, but the lesson is durable: the
efficiency an engineer should expect from a plant sized for power sits near the
square-root bound, well below the Carnot ceiling, and quoting Carnot as the target
for a working machine mistakes the zero-power limit for the design point.
\end{mastery}

\begin{mastery}
The two specific heats $c_p$ and $c_v$ were introduced as separate
measured quantities, but the first and second laws fix their difference
for any substance, not only an ideal gas. Start from the entropy as a
function of temperature and volume, $s = s(T, V)$, and from
$u = u(T, V)$. Writing $\mathrm{d}s$ and $\mathrm{d}u$ in these variables
and using $c_v = (\partial u/\partial T)_V = T(\partial s/\partial T)_V$
and $c_p = T(\partial s/\partial T)_p$, a change of variables from
$(T, V)$ to $(T, p)$ together with the Maxwell relation
$(\partial s/\partial V)_T = (\partial p/\partial T)_V$ yields the
general relation
\[
c_p - c_v = -T\,\frac{(\partial p/\partial T)_V^2}
{(\partial p/\partial V)_T}
= T V \frac{\alpha^2}{\kappa_T},
\]
where $\alpha = (1/V)(\partial V/\partial T)_p$ is the volumetric thermal
expansion coefficient and $\kappa_T = -(1/V)(\partial V/\partial p)_T$ is
the isothermal compressibility. Three readings follow at once. The
difference $c_p - c_v$ is never negative, because $\kappa_T > 0$ for any
stable substance and the other factors are squared or positive, so
$c_p \geq c_v$ universally. The difference vanishes where $\alpha = 0$,
which is why liquid water near its density maximum at $4\,\degC$ has
$c_p$ and $c_v$ almost equal, while a gas with its large expansion
coefficient has them far apart. Substituting the ideal-gas values
$\alpha = 1/T$ and $\kappa_T = 1/p$ returns $c_p - c_v = TV/(Tp/p \cdot
T) = pV/T = R$ per mole, the special case derived earlier. The general
form is what lets a property table list $c_v$ for a liquid or a solid,
where a constant-volume measurement is nearly impossible, by computing it
from the easily measured $c_p$, $\alpha$, and $\kappa_T$.
\end{mastery}

\begin{mastery}
The combined first and second laws set a ceiling on the work any state
change can deliver, and the ceiling is a state function. Define the
\engterm{flow exergy} (also called availability) of a stream relative to
an environment fixed at $T_0$ and $p_0$ as
\[
\psi = (h - h_0) - T_0 (s - s_0) + \tfrac12 V^2 + g z,
\]
where the subscript $0$ marks the property evaluated at the dead state,
the state the stream would reach in unrestricted equilibrium with the
environment. The exergy is the maximum shaft work obtainable per unit
mass as the stream is brought reversibly to the dead state. Two
properties make it the working currency of second-law analysis. First,
exergy is conserved only in a reversible process; every irreversibility
destroys exergy at the Gouy-Stodola rate $T_0 \dot{S}_{\text{gen}}$
introduced earlier, so the exergy balance for a control volume carries a
strictly non-positive destruction term where the energy balance carries
none. Second, exergy assigns a quality to energy: a joule of heat
delivered at $1500\,\si{\kelvin}$ carries far more exergy than a joule
delivered at $310\,\si{\kelvin}$, while the first law counts them
equal. The plant case study below rejects half its fuel energy at the
condenser, yet that rejected stream sits a few kelvin above the cooling
water and carries almost no exergy; the first-law loss and the
second-law loss point at different components. Volume V builds the full
exergy accounting; the point here is that $\psi$ is constructed from $h$,
$s$, and the dead state, all state functions, so the maximum work is
fixed by the endpoints even though the work actually collected depends on
the path.
\end{mastery}

\begin{mastery}
The entropy state function was introduced through $\mathrm{d}S = \delta
Q_{\text{rev}}/T$ and the inequality $\Delta S \geq \int \delta Q/T$ was
asserted. The \engterm{Clausius inequality} earns the inequality from the
Kelvin-Planck statement, and the derivation is short enough to carry. Let a
system execute an arbitrary cycle while exchanging heat $\delta Q$ with its
surroundings at the surroundings' temperature $T$. Insert between the system
and a single master reservoir at $T_0$ a reversible Carnot engine that
receives $\delta Q_0$ from the reservoir and delivers $\delta Q$ to the
system at $T$. The Carnot relation fixes the reservoir draw,
$\delta Q_0 = T_0\,\delta Q/T$, because a reversible engine transfers heat in
proportion to temperature. Over the system's full cycle the system returns to
its start, so its energy and every state function are unchanged, and the only
net effect of the combined system-plus-Carnot-engine device is that it has
drawn $\oint \delta Q_0 = T_0 \oint \delta Q/T$ from the single reservoir and
delivered an equal net work. The Kelvin-Planck statement forbids net work
extracted from a single reservoir over a cycle, so this quantity cannot be
positive,
\[
\oint \frac{\delta Q}{T} \leq 0,
\]
which is the Clausius inequality. For a reversible cycle the construction runs
backward too, forcing equality, so $\oint \delta Q_{\text{rev}}/T = 0$, which
is exactly the statement that $\delta Q_{\text{rev}}/T$ is the exact
differential of a state function, the entropy. Splitting any process from
state 1 to state 2 into the actual (possibly irreversible) leg and a
reversible return leg, and applying the inequality to the resulting cycle,
delivers $\int_1^2 \delta Q/T \leq S_2 - S_1$, the working form used
throughout. The integrating factor $1/T$ is not a guess; it is forced by the
Kelvin-Planck statement through this Carnot-engine intermediary, drawn in
Figure~\ref{fig:vol04ch01:clausius-construction}, and the same construction shows
the factor is unique up to a constant scale, which the third law then pins by
fixing the zero.
\end{mastery}

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-clausius-construction.tex}

\section{An end-to-end case study: a coal-fired steam power plant}\pathtag{standard}

The accounting of this chapter pays off when an engineer must size the
streams of a real plant and report where the fuel energy goes. We take a
coal-fired steam power plant of the kind that supplied most of the
twentieth century's electricity and carry it from the chemical energy in
the coal, through the boiler, the turbine, and the condenser, to a net
efficiency and a Sankey diagram with every stream labelled. The plant is
a good vehicle because each idea of the chapter appears in it once: the
closed-system first law in the boiler and condenser, the open-system
first law on the turbine and pump, the second-law efficiency ceiling set
by the boiler and condenser temperatures, and the entropy generation
that opens the gap between the real plant and the ceiling. The working
fluid is water, modelled here with the saturated and superheated steam
properties of standard tables \cite[Tables A-4 to A-6]{text:cengel-boles}.

\subsection{Plant specification and the cycle}

The plant generates $W_{\text{net}} = 500\,\text{MW}$ of electrical
power. It runs an ideal Rankine cycle: water leaves the condenser as
saturated liquid at $10\,\text{kPa}$ (state 1), a feed pump raises it to
the boiler pressure of $15\,\text{MPa}$ (state 2), the boiler and
superheater bring it to $540\,\degC$ (state 3), the turbine expands it
to the condenser pressure of $10\,\text{kPa}$ (state 4), and the
condenser returns it to state 1. Figure~\ref{fig:vol04ch01:rankine-ts}
draws the cycle on temperature-entropy axes. The condenser pressure of
$10\,\text{kPa}$ corresponds to a saturation temperature of about
$46\,\degC$, set by the temperature of the river or cooling tower that
carries the rejected heat; the boiler pressure and the
$540\,\degC$ steam temperature are set by the strength of the boiler
tubes and the turbine blades at temperature, the materials constraint
that has capped steam conditions for decades.

The state enthalpies, read from steam tables, are
\[
h_1 = 191.8\,\text{kJ}/\text{kg}, \quad
h_3 = 3434\,\text{kJ}/\text{kg}, \quad
h_4 = 2118\,\text{kJ}/\text{kg},
\]
where state 1 is saturated liquid at $10\,\text{kPa}$, state 3 is
superheated steam at $15\,\text{MPa}$ and $540\,\degC$, and state 4 is
the wet vapour at $10\,\text{kPa}$ reached by an isentropic expansion
from state 3 (quality $x_4 \approx 0.81$, found by matching the entropy
of state 3 to the wet-vapour entropy at $10\,\text{kPa}$). The script
\texttt{code/first\_law\_balance.py} carries the same state points and
reproduces the stream values that follow.

\subsection{The pump and the boiler}

The feed pump raises the saturated liquid from $10\,\text{kPa}$ to
$15\,\text{MPa}$. For an incompressible liquid the pump work per unit
mass is $w_{\text{pump}} = v_1 (p_2 - p_1)$, with $v_1 \approx
1.010 \times 10^{-3}\,\text{m}^3/\text{kg}$ the specific volume of
saturated liquid water,
\[
w_{\text{pump}} = 1.010 \times 10^{-3} \times (15{,}000 - 10) =
15.1\,\text{kJ}/\text{kg}.
\]
The pump enthalpy rise is small against the boiler's, so
$h_2 = h_1 + w_{\text{pump}} = 191.8 + 15.1 = 207\,\text{kJ}/\text{kg}$.
The smallness of the pump work is the structural reason a vapour power
cycle is worth building: pumping a liquid to high pressure costs little,
while expanding the vapour from that pressure returns much. A gas
compressed and then expanded over the same pressure ratio would return
nearly what it cost, which is why the working fluid is condensed to a
liquid before it is pressurised.

The boiler adds heat to carry the feedwater from state 2 to the
superheated state 3, a constant-pressure process for which the heat
added per unit mass equals the enthalpy rise,
\[
q_{\text{in}} = h_3 - h_2 = 3434 - 207 = 3227\,\text{kJ}/\text{kg}.
\]
This is the closed-system constant-pressure result $\Delta H = Q$
applied to the working fluid as it passes through the boiler tubes. It
is the largest energy stream in the plant and the one the fuel must
supply.

\subsection{The turbine and the condenser}

The turbine expands the steam from state 3 to state 4. For the ideal
(isentropic) turbine the work per unit mass is the enthalpy drop,
\[
w_{\text{turb}} = h_3 - h_4 = 3434 - 2118 = 1316\,\text{kJ}/\text{kg},
\]
the open-system first law on an adiabatic turbine with negligible
kinetic and potential terms. The condenser then rejects the heat that
carries the wet vapour at state 4 back to saturated liquid at state 1, a
constant-pressure rejection of
\[
q_{\text{out}} = h_4 - h_1 = 2118 - 191.8 = 1926\,\text{kJ}/\text{kg}.
\]
The net work per unit mass is the turbine work less the pump work,
$w_{\text{net}} = w_{\text{turb}} - w_{\text{pump}} = 1316 - 15.1 =
1301\,\text{kJ}/\text{kg}$, and the first law for the cycle as a whole
checks: $w_{\text{net}} = q_{\text{in}} - q_{\text{out}} = 3227 - 1926 =
1301\,\text{kJ}/\text{kg}$, the two routes to the net work agreeing as
the cycle integral $\oint dU = 0$ requires.

\subsection{Thermal efficiency and the mass flow}

The cycle thermal efficiency is the net work divided by the heat in,
\[
\eta_{\text{cycle}} = \frac{w_{\text{net}}}{q_{\text{in}}}
  = \frac{1301}{3227} = 0.403.
\]
The Carnot efficiency between the boiler steam temperature
($540\,\degC = 813\,\si{\kelvin}$) and the condenser temperature
($46\,\degC = 319\,\si{\kelvin}$) is $1 - 319/813 = 0.61$. The Rankine
cycle reaches two-thirds of the Carnot ceiling; the shortfall is the
heat added at temperatures below the peak (the feedwater enters the
boiler cold and is warmed across a wide range, not all at
$540\,\degC$), which lowers the average temperature of heat addition
below the Carnot reservoir temperature. Superheating and, in real
plants, feedwater reheating and regeneration claw part of this back; the
volume's chapter on cycles develops those refinements.

The steam mass flow follows from the required net power,
\[
\dot{m} = \frac{W_{\text{net}}}{w_{\text{net}}}
  = \frac{500 \times 10^3\,\text{kW}}{1301\,\text{kJ}/\text{kg}}
  = 384\,\text{kg}/\text{s}.
\]
The boiler must then supply $\dot{Q}_{\text{in}} = \dot{m}\, q_{\text{in}}
= 384 \times 3227 = 1239\,\text{MW}$, and the condenser must reject
$\dot{Q}_{\text{out}} = \dot{m}\, q_{\text{out}} = 384 \times 1926 =
740\,\text{MW}$. The condenser rejection alone exceeds the net
electrical output, which is why a large thermal plant is sited on a
river, a lake, or the coast, or carries the cooling towers that are its
most visible feature.

\subsection{From the cycle to the fuel: the overall plant balance}

The cycle efficiency of $40.3\%$ is the efficiency of the working fluid,
not of the plant. The boiler does not transfer all the fuel's energy to
the steam: some leaves up the stack as hot flue gas and uncondensed
water vapour, some radiates from the boiler casing, and some drives the
plant's own pumps, fans, and mills (the auxiliaries, or parasitic load).
A well-run pulverised-coal boiler transfers about $88\%$ of the fuel's
lower heating value to the steam \cite[ch.~10]{text:moran-shapiro-thermo};
take $\eta_{\text{boiler}} = 0.88$. The fuel energy rate is then
\[
\dot{Q}_{\text{fuel}} = \frac{\dot{Q}_{\text{in}}}{\eta_{\text{boiler}}}
  = \frac{1239}{0.88} = 1408\,\text{MW},
\]
so the stack and casing losses carry $1408 - 1239 = 169\,\text{MW}$. The
overall plant efficiency, fuel energy in to net electrical work out, is
\[
\eta_{\text{plant}} = \eta_{\text{boiler}}\,\eta_{\text{cycle}}
  = 0.88 \times 0.403 = 0.355,
\]
which the gross-to-net accounting trims further once the auxiliaries
(roughly $5\%$ of gross output in a coal plant) are subtracted, landing
the delivered efficiency near $34\%$. Modern supercritical and
ultra-supercritical coal plants push the cycle to higher steam
conditions and reach the low $40$s; the $34\%$ here is representative of
the mid-century subcritical fleet that the figures are drawn for.

To put the fuel rate in physical terms, bituminous coal has a lower
heating value near $27\,\text{MJ}/\text{kg}$. The coal burn rate is
$\dot{m}_{\text{coal}} = \dot{Q}_{\text{fuel}}/\text{LHV} =
1408 \times 10^6 / (27 \times 10^6) = 52\,\text{kg}/\text{s}$, or about
$4500\,\text{tonnes}$ a day, a unit train of coal every day or two. The
carbon in that coal leaves as roughly $9000\,\text{tonnes}$ of carbon
dioxide a day, the climate cost that the first-law accounting makes
unavoidable to state: a plant that converts a third of its fuel to
electricity sends two-thirds to the environment as heat and all of its
fuel carbon to the air.

Figure~\ref{fig:vol04ch01:plant-sankey} collects the streams into a
single energy-flow diagram drawn to scale. Of every hundred units of
fuel energy, about thirty-nine leave as net electrical work, about fifty
are rejected at the condenser, and about eleven are lost up the stack
and to radiation and auxiliaries. The diagram is the chapter's thesis in
one picture: the first law forces every band that leaves to equal the
band that arrives, and the second law forces the condenser band, the
heat rejected to the cold reservoir, to be the largest single stream.
The engineer who would raise the plant's efficiency reads the diagram
and attacks the largest avoidable loss first, which is not the
condenser (that rejection is mandated by the second law and sits at near
ambient temperature, carrying little exergy) but the irreversibility in
the boiler, where fuel burned at flame temperature transfers heat to
steam hundreds of kelvin colder. The exergy view of the mastery box
above ranks the components by $T_0 \dot{S}_{\text{gen}}$ and finds the
boiler, not the condenser, at the top of the list.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-plant-sankey.tex}

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-rankine-ts.tex}

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-polytropic.tex}

\section{A second case study: a gas turbine and where its fuel goes}\pathtag{standard}

The steam plant runs a condensing cycle and throws its largest waste
stream into a river. A gas turbine works the other way: it draws air from
the atmosphere, burns fuel in it, and exhausts the hot products straight
back to the atmosphere, carrying its largest waste stream out of a stack
at several hundred degrees. The two machines are the two halves of the
modern power station, and a combined-cycle plant bolts them together. We
take a simple-cycle industrial gas turbine of the size used for grid
peaking and for mechanical drive, carry the air from the inlet through
the compressor, combustor, and turbine to the exhaust, and report where
the fuel energy goes. Each idea of the chapter reappears: the
open-system first law on the compressor and turbine, the
constant-pressure heat addition of the combustor, the back-work ratio
that the steam plant never had to worry about, and the second-law reason
the exhaust is hot. The working fluid is air and combustion products,
modelled with the cold-air-standard assumption of constant specific
heats; the more careful variable-specific-heat treatment is left to
Volume V.

\subsection{Machine specification and the air path}

The unit delivers $W_{\text{net}} = 30\,\text{MW}$ of shaft power. Air
enters the compressor at ambient $T_1 = 290\,\si{\kelvin}$,
$p_1 = 101\,\text{kPa}$; the compressor pressure ratio is $r_p = 12$; the
combustor raises the gas to a turbine inlet temperature of
$T_3 = 1450\,\si{\kelvin}$, set by the limit the cooled first-stage
blades can survive; and the turbine expands back to atmospheric pressure.
We model the compressor and turbine with isentropic efficiencies
$\eta_{s,c} = 0.85$ and $\eta_{s,t} = 0.88$, the values a real machine of
this class carries, so the analysis lands on a believable rather than an
ideal efficiency. The fuel is natural gas with a lower heating value of
$50\,\text{MJ}/\text{kg}$. The air properties are taken as
$c_p = 1.005\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$ and $\gamma = 1.4$
throughout, the cold-air-standard simplification.

\subsection{The compressor}

The isentropic compressor exit temperature follows the pressure ratio,
$T_{2s} = T_1 r_p^{(\gamma-1)/\gamma} = 290 \times 12^{0.2857} =
590\,\si{\kelvin}$, an ideal temperature rise of $300\,\si{\kelvin}$. A
real compressor generates entropy and needs more work for the same
pressure ratio, so the actual exit is hotter,
\[
T_2 = T_1 + \frac{T_{2s} - T_1}{\eta_{s,c}}
    = 290 + \frac{300}{0.85} = 643\,\si{\kelvin}.
\]
The compressor work absorbed per unit mass of air is
$w_c = c_p (T_2 - T_1) = 1.005 \times 353 = 355\,\text{kJ}/\text{kg}$.
This is the open-system first law on an adiabatic compressor with the
kinetic and potential terms dropped; the work is supplied by the turbine
on the same shaft.

\subsection{The combustor}

The combustor adds heat at constant pressure to carry the air from the
compressor exit at $643\,\si{\kelvin}$ to the turbine inlet at
$1450\,\si{\kelvin}$. The heat added per unit mass of air is
\[
q_{\text{in}} = c_p (T_3 - T_2) = 1.005 \times (1450 - 643) =
811\,\text{kJ}/\text{kg},
\]
the constant-pressure result $\Delta h = q$ applied to the gas in the
combustor can. The compressor having pre-heated the air to
$643\,\si{\kelvin}$ is the structural reason the combustor needs less
fuel than a cold-start would suggest: the air arrives most of the way to
the firing temperature, paid for in shaft work rather than fuel.

\subsection{The turbine}

The isentropic turbine exit temperature is $T_{4s} = T_3 r_p^{-(\gamma-1)
/\gamma} = 1450/2.034 = 713\,\si{\kelvin}$, an ideal drop of
$737\,\si{\kelvin}$. The real turbine recovers only $\eta_{s,t}$ of that
drop, so the actual exit is hotter,
\[
T_4 = T_3 - \eta_{s,t}(T_3 - T_{4s}) = 1450 - 0.88 \times 737 =
801\,\si{\kelvin},
\]
and the turbine work delivered is $w_t = c_p (T_3 - T_4) = 1.005 \times
649 = 652\,\text{kJ}/\text{kg}$. The net work the shaft delivers to the
generator is the turbine work less the compressor work,
\[
w_{\text{net}} = w_t - w_c = 652 - 355 = 297\,\text{kJ}/\text{kg}.
\]
The back-work ratio is $w_c/w_t = 355/652 = 0.54$: more than half the
turbine's gross output is consumed driving its own compressor. The
irreversibilities have pushed it well above the ideal-cycle value of
$0.41$, because the real compressor costs more and the real turbine
returns less. This is the structural fragility of the gas turbine. A
small loss of turbine performance or a small rise in compressor work
eats directly into the thin margin that is net work, which is why
gas-turbine compressors and turbines are matched and washed and
monitored with a care that a steam plant's feed pump never receives.

\subsection{Thermal efficiency and the air and fuel flows}

The cycle thermal efficiency is the net work over the heat added,
\[
\eta_{\text{cycle}} = \frac{w_{\text{net}}}{q_{\text{in}}}
  = \frac{297}{811} = 0.366.
\]
The ideal cold-air-standard cycle at this pressure ratio reached
$\eta = 0.51$; the isentropic efficiencies of the real components have
cost a third of that, dropping the cycle to the high thirties, which is
representative of a good simple-cycle machine. The air mass flow follows
from the required net power,
\[
\dot{m}_a = \frac{W_{\text{net}}}{w_{\text{net}}}
  = \frac{30 \times 10^3\,\text{kW}}{297\,\text{kJ}/\text{kg}}
  = 101\,\text{kg}/\text{s},
\]
a hundred kilograms of air a second through a machine that fits in a
shipping container, the high power density that makes gas turbines the
choice for aircraft and for fast-starting grid plant. The combustor heat
rate is $\dot{Q}_{\text{in}} = \dot{m}_a q_{\text{in}} = 101 \times 811 =
81.9\,\text{MW}$, supplied by a fuel mass flow of
\[
\dot{m}_f = \frac{\dot{Q}_{\text{in}}}{\text{LHV}}
  = \frac{81.9 \times 10^6}{50 \times 10^6} = 1.64\,\text{kg}/\text{s},
\]
where we have taken the combustor itself to release nearly all the
fuel's heating value into the gas, deferring the small combustion
inefficiency to the overall balance below. The fuel-air ratio is
$\dot{m}_f/\dot{m}_a = 0.016$, a lean mixture, far below the
stoichiometric ratio of roughly $0.058$ for natural gas; the excess air
is what holds the turbine inlet temperature down to what the blades can
survive, a dilution the combustor designer manages deliberately.

\subsection{From the cycle to the fuel: the overall balance and the exhaust}

The fuel energy rate is $\dot{Q}_{\text{fuel}} = \dot{m}_f \text{LHV} =
81.9\,\text{MW}$ at the combustor; allow a small combustor and casing and
bearing loss of about $4\%$, so the energy actually reaching the gas path
is $\dot{Q}_{\text{gas}} = 0.96 \times 81.9 = 78.6\,\text{MW}$ and the
true fuel rate to deliver the rated power rises to about
$\dot{Q}_{\text{fuel}} \approx 85\,\text{MW}$. The overall efficiency,
fuel in to net shaft work out, is then
\[
\eta_{\text{plant}} = \frac{W_{\text{net}}}{\dot{Q}_{\text{fuel}}}
  = \frac{30}{85} = 0.35,
\]
in the band reported for simple-cycle machines of this size. The exhaust
carries the rest. The enthalpy flow leaving the stack relative to the
inlet air is $\dot{m}_g c_p (T_4 - T_1) = 101 \times 1.005 \times
(801 - 290) = 51.9\,\text{MW}$, the single largest stream, leaving at
$801\,\si{\kelvin}$ (about $528\,\degC$). That a stream of this size
leaves at this temperature is the whole argument for the combined cycle:
unlike the steam plant's condenser reject, which sits a few kelvin above
the cooling water and carries almost no exergy, the gas-turbine exhaust
is hot enough to raise high-pressure steam. Feeding it to a steam boiler
and turbine recovers a large part of the $52\,\text{MW}$ and lifts the
combined overall efficiency into the high fifties, the highest of any
fuel-burning machine in service.

Figure~\ref{fig:vol04ch01:gasturbine-sankey} collects the streams into a
single energy-flow diagram drawn to scale. Of every hundred units of fuel
energy, about thirty-five leave as net shaft work, about sixty-two leave
as exhaust enthalpy, and about three are lost in the combustor, casing,
and bearings. Set beside the steam-plant diagram, the lesson is that the
first law fixes the size of every band while the architecture of the
machine fixes where the largest waste band goes: into cooling water for
the condensing steam cycle, up a hot stack for the gas turbine. The
engineer reads the gas-turbine diagram and attacks the exhaust, because a
sixty-two-percent stream at $528\,\degC$ is the largest pool of
recoverable work in the plant, and the second case study ends where the
combined cycle begins.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-gasturbine-sankey.tex}

\begin{mastery}
The cold-air-standard analysis used in both the worked example and this
case study makes three simplifications that a production calculation must
eventually drop. First, it treats the specific heats as constant, when in
truth $c_p$ for air climbs from about $1.005\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$ at $290\,\si{\kelvin}$ to about $1.21\,\text{kJ}/(\text{kg}
\cdot\si{\kelvin})$ near $1450\,\si{\kelvin}$, so the hot-air-standard
treatment that evaluates $c_p$ along the path raises the combustor heat
and lowers the turbine exit temperature relative to the cold-air figures.
Second, it carries air through the whole machine, when the turbine
actually expands combustion products whose composition and gas constant
differ from air, and whose mass flow exceeds the compressor's by the fuel
added. Third, it ignores the bleed air drawn off the compressor to cool
the turbine blades, which never sees the combustor and so does no useful
expansion work, a debit of several percent on the net output. The gap
between the cold-air-standard efficiency and the measured efficiency of a
real engine is the sum of these three, plus the isentropic-efficiency
debits already carried. The cold-air-standard model earns its place not
as a final answer but as the calculation that fits on one page and gets
the architecture and the order of magnitude right, the same role the
ideal-gas law plays for a real gas: a first pass exact enough to size the
streams and locate the largest losses before the property tables come
out.
\end{mastery}

\section{A third case study: compressed-air energy storage}\pathtag{standard}

The steam plant and the gas turbine both convert fuel to work in a single
pass. A storage plant does something the first two never attempt: it
takes in work when electricity is cheap, holds it, and returns work when
electricity is dear. Compressed-air energy storage (CAES) is the
thermodynamic storage technology that the first law accounts most
cleanly, because its working fluid is air and its losses are exactly the
heat-of-compression and irreversibility terms this chapter has built up.
We carry a utility-scale CAES plant through a full charge-store-discharge
cycle, track every energy stream, and report the round-trip efficiency in
two architectures: the diabatic plant that throws the compression heat
away and burns fuel to reheat the air on discharge, and the adiabatic
plant that stores the compression heat and returns it.
Figure~\ref{fig:vol04ch01:caes-schematic} lays out both.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-caes-schematic.tex}

\subsection{Plant specification and the charge stroke}

The plant stores air in a solution-mined salt cavern of volume
$V_{\text{cav}} = 3.0 \times 10^5\,\text{m}^3$, charged between
$p_{\min} = 4.6\,\text{MPa}$ and $p_{\max} = 7.0\,\text{MPa}$, the
pressure swing a real cavern is operated across. During the off-peak
charge the plant draws air from the atmosphere at $T_0 = 290\,\si{\kelvin}$,
$p_0 = 100\,\text{kPa}$ and compresses it to cavern pressure through a
train of stages with intercoolers. We model the charge as compressing air
to a representative $p_{\text{store}} = 6.0\,\text{MPa}$, a pressure ratio
of $60$, and take the air as ideal with $c_p = 1.005\,\text{kJ}/(\text{kg}
\cdot\si{\kelvin})$, $c_v = 0.718\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$,
$R = 287\,\text{J}/(\text{kg}\cdot\si{\kelvin})$, $\gamma = 1.4$.

A single-stage adiabatic compression to a pressure ratio of $60$ would
leave the air at $T = 290 \times 60^{0.2857} = 290 \times 3.59 =
1041\,\si{\kelvin}$, far too hot for the cavern and for the seals. The
plant therefore compresses in stages with intercooling. Take three stages
of equal pressure ratio $r = 60^{1/3} = 3.91$ each, with intercoolers
returning the air to $290\,\si{\kelvin}$ between stages. The temperature
rise per stage is the same,
\[
T_{\text{out}} = 290 \times 3.91^{0.2857} = 290 \times 1.467 =
425\,\si{\kelvin},
\]
a rise of $135\,\si{\kelvin}$ per stage. The ideal (isentropic) work per
unit mass for the three-stage train is three times the per-stage work,
\[
w_{c,\text{ideal}} = 3\,c_p (T_{\text{out}} - T_0)
  = 3 \times 1.005 \times 135 = 407\,\text{kJ}/\text{kg},
\]
and a real compressor train with isentropic efficiency
$\eta_{s,c} = 0.85$ absorbs
\[
w_c = \frac{w_{c,\text{ideal}}}{\eta_{s,c}}
  = \frac{407}{0.85} = 479\,\text{kJ}/\text{kg}.
\]
The intercoolers and the aftercooler reject the heat of compression. With
the air returned to $290\,\si{\kelvin}$ after each of the three stages,
the heat rejected per unit mass is, to the cold-air-standard approximation,
\[
q_{\text{rej}} = 3\,c_p (T_{\text{out}}' - T_0),
\]
where $T_{\text{out}}'$ is the actual stage exit including the
inefficiency, $T_{\text{out}}' = T_0 + (T_{\text{out}} - T_0)/\eta_{s,c}
= 290 + 135/0.85 = 449\,\si{\kelvin}$. So $q_{\text{rej}} = 3 \times
1.005 \times 159 = 479\,\text{kJ}/\text{kg}$, equal to the work absorbed,
as the first law requires for a stream that enters and leaves the
compression-plus-cooling train at the same $290\,\si{\kelvin}$: every
joule of compression work ends as rejected heat once the air is cooled
back to ambient before storage. The cooled high-pressure air enters the
cavern at $290\,\si{\kelvin}$.

\subsection{The store and the discharge stroke}

The cavern holds the cooled air. We idealise the store as isothermal at
$290\,\si{\kelvin}$ (the surrounding salt is a large thermal reservoir on
the charge-discharge timescale), so the stored air carries no thermal
energy above ambient: in the diabatic plant the compression heat is gone,
rejected at the coolers. On discharge the air leaves the cavern at
$290\,\si{\kelvin}$ and must be reheated before expansion, because
expanding cold high-pressure air through a turbine would drop its
temperature far below ambient and risk icing and brittle fracture in the
machine. The diabatic plant burns natural gas in a combustor to raise the
air to a turbine inlet temperature $T_3 = 820\,\si{\kelvin}$ before each
expansion stage.

Model the discharge as a two-stage expansion from $6.0\,\text{MPa}$ back
to atmospheric, pressure ratio $7.75$ per stage, with reheat to
$820\,\si{\kelvin}$ before each stage and turbine isentropic efficiency
$\eta_{s,t} = 0.88$. The isentropic temperature drop per stage is
$T_3(1 - 7.75^{-0.2857}) = 820 \times (1 - 0.561) = 360\,\si{\kelvin}$,
of which the real turbine captures $\eta_{s,t}$,
\[
w_{t,\text{stage}} = \eta_{s,t}\,c_p \times 360
  = 0.88 \times 1.005 \times 360 = 318\,\text{kJ}/\text{kg},
\]
and two stages deliver $w_t = 636\,\text{kJ}/\text{kg}$. The reheat
energy supplied by the fuel is the enthalpy added before each stage. The
air enters the first reheat at $290\,\si{\kelvin}$ and the second reheat
at its exit from the first stage; taking each reheat to lift the air to
$820\,\si{\kelvin}$ from an inlet near $530\,\si{\kelvin}$ on the second
stage and from $290\,\si{\kelvin}$ on the first, the fuel heat per unit
mass is
\[
q_{\text{add}} = c_p\big[(820 - 290) + (820 - 530)\big]
  = 1.005 \times (530 + 290) = 824\,\text{kJ}/\text{kg}.
\]

\subsection{Round-trip efficiency of the diabatic plant}

The diabatic plant returns more work on discharge than it spent on
charge, because the fuel added energy on the way out. The honest
accounting counts both inputs, the charge work and the fuel heat, against
the work returned. The work ratio alone is
\[
\frac{w_t}{w_c} = \frac{636}{479} = 1.33,
\]
which would read as an efficiency above one if the fuel were ignored. It
is not ignored. The round-trip efficiency counting all energy in is
\[
\eta_{\text{rt,energy}} = \frac{w_t}{w_c + q_{\text{add}}}
  = \frac{636}{479 + 824} = \frac{636}{1303} = 0.49,
\]
and a more telling figure for a storage plant is the electricity-in to
electricity-out ratio with the fuel charged separately, the
\engterm{charging electricity efficiency}, which compares the returned
work to the charge work and reports the fuel as a parallel input. Either
way the plant is a hybrid: part storage, part gas turbine running on
pre-compressed air. The two operating CAES plants of the twentieth
century, the Huntorf plant in Germany and the McIntosh plant in Alabama,
are diabatic and report figures in this band, with the McIntosh plant
recovering some exhaust heat in a recuperator to lift the number.
Figure~\ref{fig:vol04ch01:caes-sankey} draws the diabatic balance in its
left bar.

\subsection{The adiabatic plant: storing the heat instead of buying it}

The diabatic plant's loss is visible in the charge stroke: it rejected
$479\,\text{kJ}/\text{kg}$ of perfectly good heat at the coolers, then
bought $824\,\text{kJ}/\text{kg}$ of heat back from fuel on discharge.
The adiabatic plant closes that loop. It routes the hot compressed air
through a thermal store, a packed bed of rock or a molten-salt loop, that
captures the compression heat; on discharge it sends the cold stored air
back through the same thermal store to reheat it, burning no fuel. The
dashed loop of Figure~\ref{fig:vol04ch01:caes-schematic} is that path.

Suppose the thermal store captures and returns the compression heat with
a recovery effectiveness $\varepsilon_{\text{th}} = 0.90$, so the air on
discharge is reheated to $T_0 + 0.90 \times (T_{\text{out}}' - T_0) =
290 + 0.90 \times 159 = 433\,\si{\kelvin}$ per recovered stage rather
than to the $820\,\si{\kelvin}$ that fuel provided. The expansion now runs
at a lower inlet temperature, so it returns less work, but the input is
only the charge work. With the lower turbine inlet temperature the
two-stage expansion delivers about $w_t \approx 335\,\text{kJ}/\text{kg}$,
and the round-trip efficiency, charge work in to expansion work out, is
\[
\eta_{\text{rt}} = \frac{w_t}{w_c}
  = \frac{335}{479} = 0.70,
\]
in the band that demonstration adiabatic plants target. The first law has
done the whole comparison: the diabatic plant pays twice, rejecting the
compression heat and buying reheat from fuel, while the adiabatic plant
pays once, recycling the compression heat through a store. The right bar
of Figure~\ref{fig:vol04ch01:caes-sankey} sets the two side by side. The
engineering cost of the higher efficiency is the thermal store itself, a
large insulated mass that must survive thousands of charge-discharge
thermal cycles, which is why the adiabatic plant has been slower to reach
commercial scale despite its cleaner thermodynamics.

\input{volumes/04-energy/ch01-three-laws-of-thermodynamics/figures/fig-caes-sankey.tex}

\subsection{Sizing the store and the closing balance}

The cavern's energy content sets the plant's discharge duration. The mass
of air stored between $p_{\min}$ and $p_{\max}$ at the cavern temperature
is, by the ideal-gas law,
\[
\Delta m = \frac{(p_{\max} - p_{\min}) V_{\text{cav}}}{R\,T_{\text{cav}}}
  = \frac{(7.0 - 4.6)\times 10^6 \times 3.0\times 10^5}
         {287 \times 290}
  = 8.7 \times 10^6\,\text{kg},
\]
roughly nine thousand tonnes of air cycled per charge. At the diabatic
discharge work of $636\,\text{kJ}/\text{kg}$, the stored electrical
energy delivered per cycle is $\Delta m \times w_t = 8.7\times 10^6
\times 636 \times 10^3 = 5.5\times 10^{12}\,\text{J}$, or
$1{,}540\,\text{MWh}$, enough to run a $300\,\text{MW}$ generator for
about five hours, the size that makes CAES a grid-scale resource rather
than a laboratory curiosity. The charge stroke must supply
$\Delta m \times w_c = 8.7\times 10^6 \times 479 \times 10^3 =
4.2\times 10^{12}\,\text{J} = 1{,}160\,\text{MWh}$ of off-peak
electricity, plus the cooling duty at the intercoolers, and the diabatic
discharge burns $\Delta m \times q_{\text{add}} = 8.7\times 10^6 \times
824 \times 10^3 = 7.2\times 10^{12}\,\text{J}$ of fuel heat, about
$2{,}000\,\text{MWh}$, over the five-hour discharge. The closing balance
is the chapter's thesis once more: every stream that leaves the plant
equals a stream that entered, the second law charges the round-trip
efficiency for the irreversibility in the compressor, the turbine, and
the finite-temperature heat exchange at the coolers and the store, and
the architecture decides whether the compression heat is thrown away and
bought back from fuel or recycled. The same first-law audit that sized
the steam plant and the gas turbine sizes the storage plant, and the only
new idea the storage case demanded is that the input and the output are
separated in time.

\section{Failure: perpetual motion machines and how to recognise them in proposals}\pathtag{core}

A perpetual-motion machine is a proposed device that violates one of
the laws of thermodynamics. The discipline of recognising such
proposals is part of every working engineer's tool kit, because
proposals that violate the laws continue to appear in engineering
journals, patent filings, venture pitches, and consumer products. The
proposal is sometimes naive, sometimes deliberate, and sometimes the
output of someone who has confused themselves with a long calculation.
The reader who can name the law violated and the mechanism by which
the proposal violates it is protected against most of these.

\engterm{Perpetual motion of the first kind} violates the first law:
the machine produces net work without an energy input. Any closed
machine that runs forever while delivering useful work is a
first-kind perpetual motion machine. Examples that recur in the
literature:

\begin{itemize}[leftmargin=*]
  \item The over-balanced wheel: a wheel with weights that slide
    outward on one side and inward on the other, claimed to produce a
    net torque. The wheel's centre of mass is fixed by its geometry; a
    careful balance shows the moments cancel.
  \item Magnetic motors that allegedly run on the field of permanent
    magnets without input. The energy stored in a permanent-magnet
    field is finite; once extracted, the magnet is demagnetised.
  \item Water-wheels or weight-driven devices that claim to lift more
    weight than they descend with. The work output cannot exceed the
    gravitational potential energy descended through.
  \item Devices proposed in patent filings as ``over-unity'': the
    measured output exceeds the measured input. The discrepancy is
    always traced to a measurement error, a hidden input (battery,
    grid connection, chemical fuel), or a misidentified flow.
\end{itemize}

The diagnostic is simple: draw the boundary, list every energy flow
across it, and verify the balance. A claim that the input is less
than the output points to an uncounted input, not to a law violation.

\engterm{Perpetual motion of the second kind} violates the second
law: the machine converts heat to work without rejecting heat to a
colder reservoir, or it transfers heat from a colder body to a hotter
body without input work. The class includes more sophisticated
proposals than the first kind, because the second-law violation is
not always obvious from a single balance. Examples:

\begin{itemize}[leftmargin=*]
  \item A machine that extracts work from the heat of a single
    reservoir (the ocean, the atmosphere, the warm earth). The first
    law is satisfied, but the Kelvin-Planck statement is violated: a
    cyclic device cannot extract work from a single reservoir.
  \item Maxwell's demon: a microscopic agent that sorts fast and slow
    molecules to produce a temperature difference, allowing work
    extraction. The resolution (Landauer, Bennett) is that the
    demon's measurement and memory-erasure costs are at least as much
    work as the demon can extract; the second law applies to the
    combined demon-plus-gas system.
  \item Self-cooling devices that allegedly require no input work to
    maintain a temperature difference. The Clausius statement is
    violated; the device, if real, would equilibrate to the ambient
    temperature.
  \item Vapour-compression cycles that claim a coefficient of
    performance exceeding the Carnot bound $T_c/(T_h - T_c)$ for
    their operating temperatures. The Carnot bound is a theorem of
    thermodynamics; a higher COP is a measurement error or a
    misidentification of the operating temperatures.
\end{itemize}

\engterm{Perpetual motion of the third kind} violates the third law:
the machine reaches absolute zero in a finite number of cooling
stages. The class is rarely encountered in engineering proposals; it
is included in some classifications and not in others. The third-law
violation is the operational form of the unreachability of $T = 0$.

The discipline for evaluating a proposed energy device is mechanical:

\begin{enumerate}[leftmargin=*]
  \item Draw the system boundary explicitly, on paper.
  \item List every flow of mass, work, and heat across the boundary,
    with sign convention chosen and stated.
  \item Apply the first law: does $\Delta E = Q + W$ close, with all
    inputs counted? If the claim is over-unity at this stage, the
    missing input is the diagnostic.
  \item Apply the second law: does the proposed cycle satisfy the
    Clausius and Kelvin-Planck statements? Compute the entropy
    generation and confirm it is non-negative.
  \item Compute the Carnot bound for any heat engine or refrigeration
    claim and confirm the proposed efficiency or COP lies below the
    bound.
  \item If the proposal involves chemical or electrochemical
    transformations, verify the Gibbs free-energy balance (the
    maximum useful work is $-\Delta G$ at constant $T$ and $p$).
\end{enumerate}

A proposal that survives the six steps is not a perpetual motion
machine. A proposal that fails any step is, regardless of how
sophisticated the rest of the apparatus appears. The history of
patent offices is filled with rejected over-unity claims; the U.S.
Patent and Trademark Office's standing rule that perpetual-motion
patents require a working prototype before examination is an
operational acknowledgment of the recurrence of the failure mode.

\begin{principle}[The three laws as filter]
A proposed energy device that violates the first law produces work
from no input; a proposal that violates the second law produces work
from a single reservoir or moves heat against the gradient without
input work; a proposal that violates the third law reaches absolute
zero. Every such proposal can be diagnosed by drawing the boundary,
listing the flows, and applying the three balance equations (energy,
entropy, third-law zero) in turn. The diagnostic survives every
generation of perpetual-motion proposal because the laws are
empirical, not contingent on the proposer's intent.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Instrument a household refrigerator]
The reader instruments a working household refrigerator over one week
and measures its work input, its heat rejection, and its coefficient
of performance. The deliverable is a fifteen- to twenty-five-page
report.

\textbf{Track}. Household instrumentation plus analysis.
\textbf{Hazard class}: low.

\textbf{Measurements}.

\begin{itemize}[leftmargin=*]
  \item Electrical energy consumption: a plug-in energy meter (Kill A
    Watt or equivalent) records the cumulative kWh delivered to the
    refrigerator over seven days, with the meter zeroed at the start.
  \item Internal cabinet temperature: a logging thermometer (or a
    smartphone thermometer with manual recording at $30\,\text{min}$
    intervals during waking hours) records the temperature at the
    same location for the full week.
  \item Ambient kitchen temperature: a logging thermometer at the
    refrigerator's external location records ambient over the same
    week.
  \item Condenser-coil temperature: an infrared thermometer or
    surface-contact thermometer reads the coil temperature at four
    times during the week.
  \item Cabinet contents: photograph the cabinet contents at the
    start and end of the week, noting any large additions or
    removals.
\end{itemize}

\textbf{Analysis}.

\begin{itemize}[leftmargin=*]
  \item Plot the internal temperature, ambient temperature, and
    condenser temperature over the week. Identify the
    compressor-on and compressor-off cycles in the temperature trace.
  \item Estimate the heat leak through the cabinet walls from the
    measured temperature difference, the cabinet's measured external
    dimensions, and an assumed insulation thickness and thermal
    conductivity (state the assumptions).
  \item Compute the actual coefficient of performance from $\text{COP}
    = \dot{Q}_{\text{cold}}/\dot{W}$, with $\dot{Q}_{\text{cold}}$
    estimated as the steady-state heat leak and $\dot{W}$ measured
    from the kWh meter divided by the week.
  \item Compare the measured COP to the Carnot bound for the measured
    evaporator and condenser temperatures.
  \item Identify the largest source of entropy generation in the
    cycle (the finite-temperature heat exchange at the condenser, the
    throttle valve, the compressor's mechanical losses, or the
    cabinet's heat leak), with a one-paragraph defence.
\end{itemize}

\textbf{Reflection ($1{,}500$ to $2{,}500$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
  \item What fraction of the Carnot bound did the refrigerator
    achieve? What does the gap come from?
  \item Which design changes would close the gap, and at what
    additional manufacturing or operational cost?
  \item One feature of the refrigerator's operation that the data
    revealed and the reader had not previously known.
\end{itemize}

\textbf{Deliverable}. The data log (CSV or spreadsheet), the
calculations, the report, and a one-paragraph summary suitable for
sharing with a non-engineering household member.

\textbf{Time}. One week of passive monitoring, four to eight hours of
data reduction and analysis, four to six hours of write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A closed, rigid vessel contains $2\,\text{kg}$ of air at $1\,\text{atm}$
and $20\,\degC$. The vessel is heated until the pressure reaches $3\,
\text{atm}$. Treating air as an ideal gas with $c_v = 0.718\,
\text{kJ}/(\text{kg}\cdot\si{\kelvin})$, find the heat added, the
final temperature, and the work done.
\paragraph{Solution sketch.} Rigid vessel: $V$ constant, so $W = 0$. The
ideal-gas law at fixed mass and volume gives $T_2/T_1 = p_2/p_1 = 3$, so
$T_2 = 3 \times 293\,\si{\kelvin} = 879\,\si{\kelvin}$
(i.e.\ $606\,\degC$). With $W = 0$ the first law is $Q = \Delta U =
m c_v (T_2 - T_1) = 2 \times 0.718 \times (879 - 293) =
841\,\text{kJ}$. All the added heat becomes internal energy because the
gas does no boundary work.
\end{exercise}

\begin{exercise}
A piston-cylinder device contains $0.5\,\text{kg}$ of nitrogen at
$200\,\text{kPa}$ and $300\,\si{\kelvin}$. The gas is compressed
isothermally to $800\,\text{kPa}$. Find the work done on the gas, the
heat removed, and the entropy change of the gas.
\paragraph{Solution sketch.} Nitrogen has $R = 8.314/0.028 =
296.8\,\text{J}/(\text{kg}\cdot\si{\kelvin})$. Isothermal, so
$\Delta U = 0$ and $Q = -W$. The volume ratio equals the inverse
pressure ratio, $V_2/V_1 = p_1/p_2 = 1/4$. Work on the gas:
$W = -mRT \ln(V_2/V_1) = -0.5 \times 296.8 \times 300 \times \ln(1/4)
= +61.7\,\text{kJ}$. Heat removed: $Q = -W = -61.7\,\text{kJ}$ (rejected
to the reservoir). Entropy change of the gas:
$\Delta S = mR \ln(V_2/V_1) = 0.5 \times 296.8 \times \ln(0.25) =
-205.7\,\text{J}/\si{\kelvin}$. The gas entropy falls; the reservoir's
rises by the same amount, so the total is zero for the reversible
idealisation.
\end{exercise}

\begin{exercise}
One kilogram of air ($\gamma = 1.4$, $R = 287\,\text{J}/(\text{kg}
\cdot\si{\kelvin})$) is compressed adiabatically and reversibly from
$100\,\text{kPa}$ and $300\,\si{\kelvin}$ to $500\,\text{kPa}$. Find
the final temperature, the work done on the gas, and the entropy
change.
\paragraph{Solution sketch.} Reversible adiabatic, so use the
pressure-temperature relation
$T_2 = T_1 (p_2/p_1)^{(\gamma-1)/\gamma} = 300 \times 5^{0.4/1.4} =
300 \times 5^{0.2857} = 475\,\si{\kelvin}$. With
$c_v = R/(\gamma-1) = 287/0.4 = 717.5\,\text{J}/(\text{kg}\cdot
\si{\kelvin})$, the work on the gas equals the internal-energy rise
($Q = 0$): $W = m c_v (T_2 - T_1) = 1 \times 717.5 \times (475 - 300)
= 126\,\text{kJ}$. Reversible and adiabatic means $\Delta S = 0$.
\end{exercise}

\begin{exercise}
Steam enters a turbine at $3\,\text{MPa}$ and $400\,\degC$ with
enthalpy $h_1 = 3231\,\text{kJ}/\text{kg}$ and exits at $50\,\text{kPa}$
with enthalpy $h_2 = 2403\,\text{kJ}/\text{kg}$. The mass flow rate is
$5\,\text{kg}/\text{s}$. Find the shaft power output, assuming
negligible heat loss and negligible kinetic- and potential-energy
changes.
\paragraph{Solution sketch.} Steady-state open system, adiabatic, KE
and PE negligible: the balance reduces to
$\dot{W}_s = \dot{m}(h_2 - h_1)$ for work on the CV, so the power
delivered by the turbine is $\dot{m}(h_1 - h_2) = 5 \times (3231 -
2403) = 5 \times 828 = 4140\,\text{kW} = 4.14\,\text{MW}$. The shaft
work on the CV is $-4.14\,\text{MW}$; the turbine extracts
$4.14\,\text{MW}$. The check is reproduced by
\texttt{code/first\_law\_balance.py}.
\end{exercise}

\begin{exercise}
Water enters an electric water heater at $15\,\degC$ at a flow rate of
$10\,\text{L}/\text{min}$ and leaves at $60\,\degC$. Find the
electrical power required, assuming all input power becomes heat in
the water and the water's specific heat is $4.18\,\text{kJ}/(\text{kg}
\cdot\si{\kelvin})$.
\paragraph{Solution sketch.} Mass flow:
$\dot{m} = 10\,\text{L}/\text{min} \times 1\,\text{kg}/\text{L} =
0.167\,\text{kg}/\text{s}$. Required heat rate:
$\dot{Q} = \dot{m} c \Delta T = 0.167 \times 4.18 \times (60 - 15) =
31.4\,\text{kW}$. With all electrical input converted to heat in the
water, the electrical power equals $\dot{Q} \approx 31\,\text{kW}$.
(A real domestic supply rarely delivers this on a single circuit,
which is why instantaneous heaters of this duty are uncommon.)
\end{exercise}

\begin{exercise}
A heat pump operates between an outdoor reservoir at $0\,\degC$ and an
indoor reservoir at $22\,\degC$. The Carnot coefficient of performance
for heating is $\text{COP}_h = T_h/(T_h - T_c)$. Compute the Carnot
COP. If the actual heat pump achieves $50\,\%$ of the Carnot value,
how many watts of electrical input are needed to deliver $5\,\text{kW}$
of heating to the indoor space?
\paragraph{Solution sketch.} Convert: $T_h = 295\,\si{\kelvin}$,
$T_c = 273\,\si{\kelvin}$. Carnot heating COP:
$\text{COP}_h = T_h/(T_h - T_c) = 295/22 = 13.4$. Actual COP at half
the bound: $6.7$. Electrical input to deliver $5\,\text{kW}$ of heat:
$\dot{W} = \dot{Q}_h/\text{COP} = 5000/6.7 = 746\,\text{W}$. A
resistance heater would draw the full $5\,\text{kW}$; the heat pump
delivers the same heat for roughly one-seventh of the electricity
because it moves heat rather than generating it.
\end{exercise}

\begin{exercise}
Two streams of water mix in an open vessel: $1\,\text{kg}/\text{s}$ at
$80\,\degC$ and $2\,\text{kg}/\text{s}$ at $20\,\degC$. Find the
outlet temperature and the entropy generation rate in the mixing
process. Assume negligible heat loss to the surroundings.
\paragraph{Solution sketch.} Energy balance on the adiabatic mixer:
$\dot{m}_1 c T_1 + \dot{m}_2 c T_2 = (\dot{m}_1 + \dot{m}_2) c T_{out}$,
so the outlet is the mass-weighted mean
$T_{out} = (1 \times 80 + 2 \times 20)/3 = 40\,\degC = 313\,\si{\kelvin}$.
Entropy generation (no heat crosses the boundary, so
$\dot{S}_{\text{gen}}$ is the net rise in stream entropy), with
$c = 4.18\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$ and absolute
temperatures $T_1 = 353$, $T_2 = 293$, $T_{out} = 313\,\si{\kelvin}$:
\[
\dot{S}_{\text{gen}} = \dot{m}_1 c \ln\frac{T_{out}}{T_1} +
\dot{m}_2 c \ln\frac{T_{out}}{T_2}
= 4.18\!\left(1\ln\tfrac{313}{353} + 2\ln\tfrac{313}{293}\right).
\]
The first term is $-0.503$, the second $+0.552$, summing to
$+0.049\,\text{kW}/\si{\kelvin} = 49\,\text{W}/\si{\kelvin}$. The
generation is positive, as the second law requires for an irreversible
mixing.
\end{exercise}

\begin{exercise}
A rigid tank of volume $0.1\,\text{m}^3$ initially contains air at
$200\,\text{kPa}$ and $25\,\degC$. A valve is opened and air leaks to
the atmosphere ($100\,\text{kPa}$, $25\,\degC$) until the tank reaches
atmospheric pressure. Find the entropy change of the air that
remained in the tank (assume the leak is slow and isothermal).
\paragraph{Solution sketch.} Track the gas that stays behind. Slowly
and isothermally, it expands to fill the whole tank as the rest leaks
out; its own pressure falls from $200$ to $100\,\text{kPa}$, which for a
control mass at constant $T$ is an isothermal expansion with
$\Delta S = m_{\text{rem}} R \ln(p_1/p_2)$. The final mass in the tank
is $m_{\text{rem}} = p_2 V/(RT) = 100{,}000 \times 0.1/(287 \times
298) = 0.117\,\text{kg}$, so $\Delta S = 0.117 \times 287 \times
\ln 2 = +23.2\,\text{J}/\si{\kelvin}$. The entropy of the retained gas
rises because it expands; the heat to keep it isothermal came from the
surroundings.
\end{exercise}

\begin{exercise}
Compute the entropy change when $1\,\text{mol}$ of nitrogen at
$1\,\text{atm}$ and $25\,\degC$ is mixed with $1\,\text{mol}$ of oxygen
at the same conditions, with the final mixture occupying twice the
initial volume of each. State whether the process is reversible.
\paragraph{Solution sketch.} Each gas doubles the volume available to
it, so the mixing entropy is the sum of two isothermal expansions:
$\Delta S = -R(n_{N_2}\ln x_{N_2} + n_{O_2}\ln x_{O_2})$ with
$x_{N_2} = x_{O_2} = 0.5$. Then $\Delta S = -R(1\ln 0.5 + 1\ln 0.5) =
2R\ln 2 = 2 \times 8.314 \times 0.693 = 11.5\,\text{J}/\si{\kelvin}$.
The process is irreversible: $\Delta S > 0$ for an isolated system, and
unmixing would require external work.
\end{exercise}

\begin{exercise}
A solid copper block ($m = 2\,\text{kg}$, $c = 385\,\text{J}/(\text{kg}
\cdot\si{\kelvin})$) at $100\,\degC$ is dropped into a large lake at
$15\,\degC$. Find the entropy change of the copper, the entropy
change of the lake, and the total entropy change.
\paragraph{Solution sketch.} Absolute temperatures: copper $373\to
288\,\si{\kelvin}$, lake fixed at $288\,\si{\kelvin}$. Copper:
$\Delta S_{Cu} = m c \ln(T_2/T_1) = 2 \times 385 \times \ln(288/373) =
-199\,\text{J}/\si{\kelvin}$. Heat dumped into the lake:
$Q = m c (T_1 - T_2) = 2 \times 385 \times 85 = 65.5\,\text{kJ}$,
received reversibly at the lake's fixed temperature, so
$\Delta S_{\text{lake}} = +Q/T_{\text{lake}} = 65{,}450/288 =
+227\,\text{J}/\si{\kelvin}$. Total:
$\Delta S_{\text{tot}} = -199 + 227 = +28\,\text{J}/\si{\kelvin} > 0$.
The drop is irreversible: heat crossed a finite temperature difference.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Starting from the first law for a closed system and the definition of
enthalpy $H = U + pV$, derive $\Delta H = Q$ for a closed system
undergoing a constant-pressure quasi-static process.
\paragraph{Solution sketch.} First law: $\Delta U = Q + W$ with boundary
work $W = -\int p\,dV = -p\,\Delta V$ at constant $p$. So $Q = \Delta U
+ p\,\Delta V$. At constant pressure, $\Delta(pV) = p\,\Delta V$, hence
$Q = \Delta U + \Delta(pV) = \Delta(U + pV) = \Delta H$. The enthalpy
change equals the heat exchanged precisely because the $p\,\Delta V$
flow-work term is absorbed into the definition of $H$.
\end{exercise}

\begin{exercise}
Derive the relation $c_p - c_v = R$ for an ideal gas from the
definitions of $c_p$ and $c_v$ and the ideal-gas equation of state
$pV = nRT$.
\paragraph{Solution sketch.} Per mole, $H = U + pV = U + RT$ for an
ideal gas. Differentiate with respect to $T$:
$dH/dT = dU/dT + R$. The constant-pressure and constant-volume specific
heats are $c_p = (dH/dT)_p$ and $c_v = (dU/dT)_v$; for an ideal gas $U$
depends on $T$ alone, so both derivatives are ordinary, giving
$c_p = c_v + R$, i.e.\ $c_p - c_v = R$. The relation says the extra heat
a constant-pressure process needs over a constant-volume one, per
degree, is exactly the work $R\,dT$ done in expanding.
\end{exercise}

\begin{exercise}
For an ideal gas with constant $c_v$, derive the expression $\Delta S
= n c_v \ln(T_2/T_1) + nR \ln(V_2/V_1)$ from the combined first and
second laws $dU = T\,dS - p\,dV$.
\paragraph{Solution sketch.} Solve the combined law for $dS$:
$dS = (dU + p\,dV)/T$. For an ideal gas $dU = n c_v\,dT$ and $p/T =
nR/V$, so $dS = n c_v\,dT/T + nR\,dV/V$. Integrate each term from state
1 to state 2 with $c_v$ constant:
$\Delta S = n c_v \ln(T_2/T_1) + nR \ln(V_2/V_1)$. The temperature
term is the entropy of heating; the volume term is the entropy of
expansion. The $(T,p)$ form follows by substituting $V \propto T/p$.
\end{exercise}

\begin{exercise}
Derive the Maxwell relation $(\partial S/\partial V)_T = (\partial p/
\partial T)_V$ from the differential of the Helmholtz free energy
$dA = -S\,dT - p\,dV$.
\paragraph{Solution sketch.} From $dA = -S\,dT - p\,dV$ read off the
first partials: $(\partial A/\partial T)_V = -S$ and $(\partial A/
\partial V)_T = -p$. Because $A$ is a state function, its mixed second
partials are equal: $\partial^2 A/\partial V\,\partial T = \partial^2 A/
\partial T\,\partial V$. Differentiating $-S$ with respect to $V$ and
$-p$ with respect to $T$ and equating gives $-(\partial S/\partial V)_T
= -(\partial p/\partial T)_V$, hence $(\partial S/\partial V)_T =
(\partial p/\partial T)_V$. The Maxwell relation trades an entropy
derivative for one obtainable from $pVT$ data.
\end{exercise}

\begin{exercise}
Show that for an ideal gas undergoing a reversible adiabatic process,
$pV^\gamma$ is constant, where $\gamma = c_p/c_v$.
\paragraph{Solution sketch.} Adiabatic and reversible: $\delta Q = 0$,
so the first law per mole gives $c_v\,dT = -p\,dV$. Differentiate the
ideal-gas law $pV = RT$: $p\,dV + V\,dp = R\,dT$. Eliminate $dT$ using
$dT = -p\,dV/c_v$ and use $R = c_p - c_v$:
$p\,dV + V\,dp = -(c_p - c_v)\,p\,dV/c_v$. Collect terms:
$V\,dp = -(c_p/c_v)\,p\,dV$, i.e.\ $dp/p = -\gamma\,dV/V$. Integrating,
$\ln p + \gamma \ln V = \text{const}$, so $pV^\gamma$ is constant.
\end{exercise}

\begin{exercise}
Derive the steady-state open-system first law from the closed-system
first law applied to a control mass passing through a control volume.
Identify the flow-work term $pv$ that combines with $u$ to form $h$.
\paragraph{Solution sketch.} Take a fixed mass that enters the CV
through the inlet and leaves through the outlet. Pushing a slug of mass
$\delta m$ in at pressure $p_1$ requires the upstream fluid to do work
$p_1 v_1\,\delta m$ on it; pushing the slug out at $p_2$ has the CV do
work $p_2 v_2\,\delta m$ on the downstream fluid. The closed-system
balance for the control mass is $\Delta U = Q + W_{\text{shaft}} +
(p_1 v_1 - p_2 v_2)\delta m$. Grouping the flow-work terms with the
internal energy, $u + pv = h$, gives, at steady state with rate forms,
$0 = \dot{Q} + \dot{W}_s + \dot{m}(h_1 - h_2)$, plus the kinetic and
potential terms if retained. The flow work $pv$ is what promotes $u$ to
$h$ for any stream crossing a boundary at non-zero pressure.
\end{exercise}

\begin{exercise}
Show that the Clausius statement and the Kelvin-Planck statement of
the second law are equivalent: a device that violates either could be
combined with a Carnot engine to produce a device that violates the
other.
\paragraph{Solution sketch.} Show the contrapositive both ways.
Suppose a Kelvin-Planck violator exists: a device that takes $Q$ from
the hot reservoir and produces work $W = Q$ with no rejection. Feed
that work to a Carnot refrigerator pumping $Q_c$ from the cold
reservoir and delivering $Q_c + W$ to the hot one. The combined device
moves net heat $Q_c$ from cold to hot with no external work, a Clausius
violation. Conversely, a Clausius violator (heat $Q_c$ from cold to hot
unaided) can be paired with a Carnot engine that takes $Q_c + W$ from
the hot reservoir, returns exactly $Q_c$ to the cold one, and nets work
$W$; the cold reservoir is unchanged, so the pair extracts work from a
single (hot) reservoir, a Kelvin-Planck violation. Each statement
implies the other.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the heat capacity of one kilogram of dry air at constant
pressure at $300\,\si{\kelvin}$, and the heat required to warm a
$30\,\text{m}^3$ room from $15\,\degC$ to $22\,\degC$ without
infiltration.
\paragraph{Reference estimate.} Air $c_p \approx
1.0\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$ near room temperature
(from $c_p = \gamma R/((\gamma-1)M) = 1.4 \times 8.314/(0.4 \times
0.029) \approx 1.0$). Room air mass: density $\approx
1.2\,\text{kg}/\text{m}^3$ gives $m = 1.2 \times 30 = 36\,\text{kg}$.
Heat to warm by $7\,\si{\kelvin}$: $Q = m c_p \Delta T = 36 \times 1.0
\times 7 = 252\,\text{kJ}$, about $0.07\,\text{kWh}$. A few minutes of a
$1\,\text{kW}$ heater, ignoring the much larger task of warming the
walls and contents.
\end{exercise}

\begin{exercise}
Estimate the work required to compress one cubic metre of air from
$1\,\text{atm}$ to $7\,\text{atm}$ isothermally at $300\,\si{\kelvin}$.
Compare to the work for the same compression done adiabatically.
\paragraph{Reference estimate.} Moles in $1\,\text{m}^3$ at $1\,\text{atm}$,
$300\,\si{\kelvin}$: $n = pV/RT = 101{,}325 \times 1/(8.314 \times 300)
= 40.6\,\text{mol}$. Isothermal work on the gas:
$W = nRT\ln(p_2/p_1) = 40.6 \times 8.314 \times 300 \times \ln 7 =
197\,\text{kJ}$. Adiabatic: $T_2 = 300 \times 7^{0.4/1.4} = 523\,
\si{\kelvin}$, and $W = n c_v \Delta T = 40.6 \times (2.5 \times 8.314)
\times 223 = 188\,\text{kJ}$ of internal-energy rise. The adiabatic
work is comparable but leaves the gas hot; if the gas is then cooled
back to $300\,\si{\kelvin}$ at constant pressure the total energy bill
exceeds the isothermal case. Isothermal compression is the cheaper
route, which is why large compressors are intercooled.
\end{exercise}

\begin{exercise}
Estimate the entropy change of the universe when a kettle boils
$1\,\text{L}$ of water from $20\,\degC$ to $100\,\degC$ on an
electric burner that operates at $700\,\degC$.
\paragraph{Reference estimate.} Heat into the water:
$Q = m c \Delta T = 1 \times 4.18 \times 80 = 334\,\text{kJ}$. Water
entropy gain (heating, not a phase change):
$\Delta S_{\text{water}} = mc\ln(T_2/T_1) = 4180\ln(373/293) =
+1010\,\text{J}/\si{\kelvin}$. The burner surface at $973\,\si{\kelvin}$
gives up the same $334\,\text{kJ}$, losing
$\Delta S_{\text{burner}} = -334{,}000/973 = -343\,\text{J}/
\si{\kelvin}$. Universe:
$\Delta S \approx 1010 - 343 = +670\,\text{J}/\si{\kelvin}$, positive
and dominated by delivering heat across the large burner-to-water
temperature gap.
\end{exercise}

\begin{exercise}
Estimate the order of magnitude of the daily food-energy intake of an
adult human in joules, and compare to the daily energy consumption of
a household refrigerator.
\paragraph{Reference estimate.} A daily intake of $2000\,\text{kcal}$
is $2000 \times 4184 = 8.4\,\text{MJ}$, equivalent to $2.3\,\text{kWh}$.
A refrigerator drawing $\sim 1\,\text{kWh}/\text{day}$ (label figures
of $\sim 350\,\text{kWh}/\text{year}$) uses roughly half the energy a
person eats. Averaged over the day the human metabolises at about
$8.4\,\text{MJ}/86{,}400\,\text{s} \approx 100\,\text{W}$, the same
order as the refrigerator's average electrical draw.
\end{exercise}

\begin{exercise}
Estimate the maximum work that can be obtained from $1\,\text{kg}$ of
hot tea at $80\,\degC$ delivered to a $20\,\degC$ environment, if the
tea is brought to ambient temperature reversibly. State the
calculation as an entropy bound.
\paragraph{Reference estimate.} The maximum work is the exergy of the
hot tea relative to the $T_0 = 293\,\si{\kelvin}$ environment,
$W_{\max} = mc[(T_1 - T_0) - T_0\ln(T_1/T_0)]$ with $T_1 = 353\,
\si{\kelvin}$. Numerically
$W_{\max} = 4180[(353-293) - 293\ln(353/293)] = 4180[60 - 52.6] =
22.6\,\text{kJ}$. The bracketed form is the entropy bound: the
$60\,\si{\kelvin}$ of enthalpy drop is discounted by $T_0$ times the
entropy the tea sheds, $T_0\,\Delta S_{\text{tea}}$, which cannot be
recovered as work. The heat content released is $mc\Delta T =
251\,\text{kJ}$, so only about $9\,\%$ of it is available as work
against this environment.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a $p$-$V$ diagram plotter for an ideal-gas cycle composed of
two isothermal and two adiabatic segments (a Carnot cycle). Plot the
cycle and compute the net work and the thermal efficiency. Verify
that the computed efficiency matches the Carnot bound $1 - T_c/T_h$.
\paragraph{Solution sketch.} Fix $T_h$, $T_c$, $\gamma$, and the start
state. Generate each leg parametrically: isotherms as $p = nRT/V$ over
the chosen $V$ range, adiabats as $p = p_{\text{start}}(V_{\text{start}}
/V)^\gamma$, matching the adiabatic relation $T V^{\gamma-1} =
\text{const}$ to set the corner volumes. Net work is the enclosed area
$\oint p\,dV$, evaluated by the trapezoid rule on the plotted points,
or equivalently from $Q_h - Q_c$ with $Q_h = nRT_h\ln(V_2/V_1)$.
Efficiency $\eta = W_{\text{net}}/Q_h$ should equal $1 - T_c/T_h$ to
plotting precision. The reference implementation is
\texttt{code/carnot\_cycle.py}, which asserts the match to machine
precision.
\end{exercise}

\begin{exercise}
Simulate the mixing of two ideal gases in a $1\,\text{L}$ container
divided by a partition, using a stochastic particle simulation
($10^4$ particles, periodic boundary conditions, elastic collisions).
Plot the time-evolution of the species concentrations on each side and
the total entropy (computed from the bin populations) as a function
of time.
\paragraph{Solution sketch.} Initialise species $A$ on the left and
$B$ on the right; advance positions with elastic collisions (or, more
simply, a diffusive random hop between bins). At each step estimate the
entropy from the bin compositions using $-R\sum_i x_i\ln x_i$ summed
over bins, weighted by occupancy. The concentrations equalise across
the partition and the entropy rises monotonically (apart from
stochastic noise) toward $R\ln 2$ per mole for an equimolar start. The
curve is the kinetic face of the mixing-entropy formula of the worked
examples; \texttt{code/entropy\_of\_mixing.py} carries a minimal
implementation and prints the limiting value.
\end{exercise}

\begin{exercise}
Implement a steady-state open-system energy-balance solver: given
inlet conditions, outlet conditions, work, and heat for any chosen
control volume, return the residual of the first law and flag any
inconsistency. Verify on the steam turbine of the Calculation
exercises.
\paragraph{Solution sketch.} Implement the residual
$r = \dot{Q} + \dot{W}_s + \dot{m}(h_1 - h_2 + \tfrac12(V_1^2 - V_2^2)
+ g(z_1 - z_2))$ and flag $|r|$ above a tolerance. For the steam
turbine ($\dot{m} = 5$, $h_1 = 3231$, $h_2 = 2403\,\text{kJ}/\text{kg}$,
adiabatic, KE and PE negligible), solving for $\dot{W}_s$ returns
$-4.14\,\text{MW}$ and the residual is zero. The reference
implementation \texttt{code/first\_law\_balance.py} carries both the
residual and the solve, with an assertion on the turbine case.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify the size of an electric water heater (in kW) to deliver
$50\,\text{L}/\text{min}$ at $45\,\degC$ to a residential shower,
with inlet water at $10\,\degC$. State the assumptions and the
energy efficiency assumed.
\paragraph{Solution sketch.} Flow $\dot{m} = 50/60 =
0.83\,\text{kg}/\text{s}$, rise $\Delta T = 35\,\si{\kelvin}$. Heat
rate $\dot{Q} = \dot{m}c\Delta T = 0.83 \times 4.18 \times 35 =
122\,\text{kW}$. Resistance heating converts nearly all the electrical
input to heat at the element, so specify roughly $120$ to
$130\,\text{kW}$. (This is
implausibly large for a domestic supply, which is the engineering
point: a tankless heater at this flow and rise needs a commercial feed,
so real showers use storage tanks or lower flow.) Assumptions: steady
flow, no distribution loss, water $c = 4.18\,\text{kJ}/(\text{kg}\cdot
\si{\kelvin})$.
\end{exercise}

\begin{exercise}
Design a passive solar water heating panel that delivers
$200\,\text{L}/\text{day}$ at $50\,\degC$ in a climate where the
inlet water is at $15\,\degC$ and the average solar irradiance is
$200\,\text{W}/\text{m}^2$ over the day. Compute the required
collector area and state the heat-loss assumption.
\paragraph{Solution sketch.} Daily heat demand:
$Q = 200 \times 4.18 \times 35 = 29.3\,\text{MJ}/\text{day}$. Daily
insolation per square metre: $200\,\text{W}/\text{m}^2 \times
86{,}400\,\text{s} = 17.3\,\text{MJ}/(\text{m}^2\cdot\text{day})$. With
a perfect collector the area is $29.3/17.3 = 1.7\,\text{m}^2$; with a
realistic collector efficiency of $50\,\%$ (optical and convective
losses), $A \approx 3.4\,\text{m}^2$. State the heat-loss assumption
explicitly: the $50\,\%$ lumps glazing reflection, absorber
re-radiation, and convective loss to ambient; a real datasheet gives
this as the collector's efficiency curve against $(T_{\text{plate}} -
T_{\text{amb}})/G$.
\end{exercise}

\begin{exercise}
Specify a heat-recovery ventilation unit for a residential building
where the indoor temperature is $22\,\degC$, the outdoor temperature
is $-5\,\degC$, and the ventilation rate is $100\,\text{m}^3/\text{h}$.
State the heat-recovery effectiveness target and compute the
auxiliary heating power required after recovery.
\paragraph{Solution sketch.} Mass flow:
$\dot{m} = 1.2 \times 100/3600 = 0.033\,\text{kg}/\text{s}$. Without
recovery, warming the incoming air across $\Delta T = 27\,\si{\kelvin}$
needs $\dot{m}c_p\Delta T = 0.033 \times 1005 \times 27 =
905\,\text{W}$. Set an effectiveness target $\varepsilon = 0.8$
(typical for a good plate or rotary core); the supply air leaves the
exchanger at $-5 + 0.8 \times 27 = 16.6\,\degC$. The auxiliary heater
then supplies $\dot{m}c_p(22 - 16.6) = 0.033 \times 1005 \times 5.4 =
181\,\text{W}$, a fivefold reduction. The recovery does not violate the
second law: it transfers heat from the warm exhaust to the cold supply
across a finite gradient, generating entropy, and never lifts the
supply above the exhaust temperature.
\end{exercise}

\begin{exercise}
A household kettle holds $1.7\,\text{L}$ of water and boils a full
load from $20\,\degC$ to $100\,\degC$. Specify the electrical power
rating to bring the water to a boil in not more than three minutes,
and state the efficiency assumption.
\paragraph{Solution sketch.} Heat to boil:
$Q = mc\Delta T = 1.7 \times 4.18 \times 80 = 569\,\text{kJ}$. Time
limit $t = 180\,\text{s}$, so the heat rate is $569/180 =
3.16\,\text{kW}$ at perfect efficiency. Allow for heat lost to the
kettle body and steam by assuming $85\,\%$ efficiency, giving a rating
of $3.16/0.85 = 3.7\,\text{kW}$. A standard domestic kettle is rated
$2$ to $3\,\text{kW}$, so a full $1.7\,\text{L}$ load on a
$2.2\,\text{kW}$ element takes closer to five minutes; meeting the
three-minute target needs a $3\,\text{kW}$-plus circuit.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A household refrigerator's measured energy consumption (from the
appliance label) is $400\,\text{kWh}/\text{year}$. The cabinet
volume is $400\,\text{L}$, the average ambient temperature is
$22\,\degC$, and the internal compartment is $4\,\degC$. Compute the
implied average heat leak in watts and infer the cycle's effective
COP. Identify the largest term in the energy balance.
\paragraph{Solution sketch.} Average electrical power:
$400\,\text{kWh}/\text{year} = 400{,}000/8760 = 45.7\,\text{W}$.
Conduction leak for a $400\,\text{L}$ cabinet (surface area
$\sim 4.5\,\text{m}^2$, foam $5\,\si{\centi\meter}$, $k =
0.025\,\text{W}/(\text{m}\cdot\si{\kelvin})$, $\Delta T = 18\,
\si{\kelvin}$): $\dot{Q}_{\text{leak}} = kA\Delta T/L = 0.025 \times
4.5 \times 18/0.05 = 40\,\text{W}$. Treating that as the entire cold
load gives an implied $\text{COP} = 40/45.7 = 0.9$, which is below one
and therefore not physical for a working cycle. The diagnostic
conclusion: the steady conduction leak is not the largest term. Door
openings, defrost heaters, and the fan and control electronics add to
the load and raise the true cold-side heat above the conduction
estimate, so the real cycle COP exceeds one while the label energy
still resolves to $45.7\,\text{W}$ average. The largest term in a
typical cabinet balance is the combined infiltration-plus-defrost load,
not the wall conduction.
\end{exercise}

\begin{exercise}
A consumer-grade thermos flask holds $1\,\text{L}$ of coffee
initially at $90\,\degC$ in a $20\,\degC$ ambient. After eight hours
the coffee is at $60\,\degC$. Estimate the effective $UA$ value of
the flask (heat-transfer coefficient times area) and the
corresponding overall thermal conductance in $\text{W}/\si{\kelvin}$.
\paragraph{Solution sketch.} Lumped cooling:
$T(t) - T_a = (T_0 - T_a)e^{-UA\,t/(mc)}$. With $T_0 = 90$, $T_a = 20$,
$T(8\,\text{h}) = 60\,\degC$, $m = 1\,\text{kg}$, $c =
4180\,\text{J}/(\text{kg}\cdot\si{\kelvin})$, solve for $UA$:
\[
UA = -\frac{mc}{t}\ln\frac{T(t) - T_a}{T_0 - T_a}
= -\frac{4180}{28{,}800}\ln\frac{40}{70} = 0.081\,\text{W}/\si{\kelvin}.
\]
The overall thermal conductance is about $0.08\,\text{W}/\si{\kelvin}$:
a vacuum flask leaks under a tenth of a watt per kelvin, which is why
it holds heat for hours where a ceramic mug ($UA$ of order
$1\,\text{W}/\si{\kelvin}$) cools in minutes.
\end{exercise}

\begin{exercise}
An espresso machine's heating element is rated $1200\,\text{W}$ and
heats $50\,\text{mL}$ of water from $20\,\degC$ to $90\,\degC$ in
$15\,\text{s}$. Compute the apparent efficiency of the heating
process. Identify where the missing energy is going.
\paragraph{Solution sketch.} Heat into the water:
$Q = mc\Delta T = 0.05 \times 4180 \times 70 = 14{,}630\,\text{J}$.
Electrical energy supplied: $1200 \times 15 = 18{,}000\,\text{J}$.
Apparent efficiency $= 14{,}630/18{,}000 = 0.81$. The missing $19\,\%$
goes into heating the boiler block and group head (their thermal mass
soaks up energy that never reaches this shot), into conduction and
convection loss from the hot metal, and into the warm-up transient
before the element reaches steady output. A machine left on to
pre-heat amortises the block-heating term across many shots, raising
the per-shot efficiency.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A 2024 patent application claims a device that uses a permanent
magnet to drive a small generator that recharges its own input
battery while delivering net power to a load. Identify the law
violated, the boundary on which the violation appears, and the
single measurement that would detect it.
\paragraph{Solution sketch.} This is perpetual motion of the first
kind: it claims net power out while recharging its own source, so a
boundary drawn around the whole sealed device shows work leaving with
no energy entering, in violation of $\Delta E = Q + W$. The permanent
magnet stores a finite field energy that demagnetises once extracted;
it is not a perpetual source. The single detecting measurement: meter
the battery's state of charge (or its terminal energy) over a run with
the external load connected. If the device were real the battery would
not deplete; in every such case it does. Equivalently, integrate net
electrical power across the device boundary over time and confirm it is
not positive.
\end{exercise}

\begin{exercise}
A venture pitch describes an ``ambient energy harvester'' that
extracts useful electrical work from the thermal motion of air at
room temperature without any temperature gradient or other driving
potential. Identify the second-law violation, draw the equivalent
Kelvin-Planck construction, and write a single sentence rejecting the
proposal.
\paragraph{Solution sketch.} The harvester claims to take heat from a
single reservoir (room-temperature air, with no second colder
reservoir) and turn it into work in a cycle. That is exactly the
Kelvin-Planck construction: a cyclic device drawing heat $Q$ from one
reservoir at $T$ and delivering work $W = Q$ with no rejection. The
first law is satisfied; the second law is not. Rejection sentence: ``A
cyclic engine cannot convert heat from a single uniform-temperature
reservoir into work, so without a colder sink or another driving
potential the device produces no net power and is rejected on the
Kelvin-Planck statement.''
\end{exercise}

\begin{exercise}
A consumer product is sold as a refrigerator that operates with no
external power input. The cabinet is internally cooled by an
``advanced thermoelectric'' core. Identify the implied claim about
heat flow, the law it violates, and the empirical test that would
detect the violation.
\paragraph{Solution sketch.} A refrigerator with no power input claims
to move heat from the cold cabinet to the warmer room with no
compensating work, which is the Clausius statement violated outright:
heat does not pass from a colder body to a hotter one of itself. A
thermoelectric (Peltier) core still needs electrical input; called on
to run with none, it cannot maintain a temperature difference. The
empirical test: seal the unit with no electrical or chemical input and
log the internal temperature. By the second law it equilibrates to
ambient; a genuine self-cooling claim would hold the cabinet below room
temperature indefinitely, which never happens. Measuring zero input
power while the cabinet stays cold is the contradiction.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A colleague proposes to compute the efficiency of a coal-fired power
plant as $\eta = W_{\text{net}}/Q_{\text{in}}$ but to use the heating
value of coal rather than the enthalpy of the combustion products as
$Q_{\text{in}}$. Write a one-paragraph reply identifying the
inconsistency and naming the correct quantity.
\paragraph{Model-answer sketch.} The numerator and denominator must
refer to the same accounting boundary. ``Heating value'' is a defined
quantity (the enthalpy of combustion to a stated reference state, with
the higher heating value assuming the product water condenses and the
lower heating value assuming it leaves as vapour), so the issue is
consistency rather than the wrong kind of number. The correct
denominator is the heat actually released into the cycle's working
fluid, which for a plant venting hot flue gas and uncondensed water is
the lower heating value times the fuel rate, not the higher. Mixing the
higher heating value into $Q_{\text{in}}$ while the cycle never recovers
the condensation enthalpy understates the efficiency; state which
heating value is used and apply it consistently to both the fuel input
and any recovered heat.
\end{exercise}

\begin{exercise}
A student argues that the second law is "really just" the first law
because both express conservation. Write a one-paragraph reply
distinguishing the two laws by what they conserve and by the
asymmetry of allowed processes.
\paragraph{Model-answer sketch.} The first law conserves energy: the
total energy of an isolated system is constant, and any process keeps
the books balanced. The second law conserves nothing; it asserts an
inequality, that the entropy of an isolated system cannot decrease.
Both directions of a heat transfer between hot and cold bodies conserve
energy, so the first law permits the cold body to warm the hot one. The
second law forbids it, because only one direction increases total
entropy. The first law tells what is allowed by the energy balance; the
second law selects which of the energy-conserving processes actually
happen and sets the maximum work obtainable. They are different
statements about different quantities, not two versions of
conservation.
\end{exercise}

\begin{exercise}
A facility manager argues that the refrigerator should be left fully
loaded because a fuller refrigerator uses less energy. Identify
whether the claim is correct and state the dominant mechanism.
\paragraph{Model-answer sketch.} The claim is partly correct but for
the right reason, and only in a particular operating pattern. At steady
state the compressor work tracks the cabinet heat load, which is set by
the wall conduction and the door-opening losses, not by how full the
cabinet is; a full and an empty refrigerator at the same temperature
leak the same heat through the same walls. The dominant mechanism that
makes a fuller unit appear more efficient is thermal mass: the food and
liquids hold cold, so each door opening warms the air but the bulk
contents recover the temperature with a smaller compressor duty cycle,
and the cabinet rides through outages and defrosts with less swing.
Filling with water bottles to add mass helps a frequently-opened unit;
it does nothing for a sealed one.
\end{exercise}
