\chapter{Electromagnetic waves and antennas}
\label{vol04:ch10}

\epigraph{The experiments described in this paper appear to me to be
of importance as a means of investigating the existence and properties
of those electric oscillations which Maxwell's theory has predicted
should take place when an electric system is suddenly disturbed.}
{Heinrich Hertz, \emph{Electric Waves}, English trans.\ D.\ E.\ Jones
(London: Macmillan, 1893), Introduction \cite[p.~1]{hist:hertz1893}.}

\chapmeta{Half-life: durable. Archetypes: transport, interface. Exercise target: 31. Project track: physical build plus instrumented measurement (hybrid). Hazard class: low (low-power RF, $\le 10\,\si{\milli\watt}$ transmit). Reader path default: standard, with sections explicitly tagged. Named cases: none directly; impedance-mismatch and standing-wave failure modes described by mechanism.}

\noindent
Maxwell's equations in free space imply the wave equation for the
electric and magnetic fields. The wave propagates at the speed of
light, carries energy via the Poynting vector, and reflects, refracts,
and diffracts at material boundaries. The engineering consequences are
transmission lines (waveguides for one or two conductors), waveguides
(hollow metal pipes that guide the field), antennas (transducers
between guided waves and free-space waves), and the failure modes of
mismatch and standing waves that haunt RF design.

\section{The wave equation from Maxwell}\pathtag{core}

In free space ($\rho = 0$, $\vect{J} = \vect{0}$) the four Maxwell
equations reduce to
\[
\divg \vect{E} = 0, \quad \divg \vect{B} = 0,
\]
\[
\curl \vect{E} = -\pdv{\vect{B}}{t}, \quad \curl \vect{B} =
\mu_{0}\varepsilon_{0}\pdv{\vect{E}}{t}.
\]
Taking the curl of Faraday's law and substituting Amp\`ere-Maxwell,
together with the identity $\curl(\curl \vect{E}) = \grad(\divg
\vect{E}) - \laplacian \vect{E}$ and the vacuum condition
$\divg \vect{E} = 0$, gives
\[
\laplacian \vect{E} - \frac{1}{c^{2}}\frac{\partial^{2}\vect{E}}{
\partial t^{2}} = \vect{0}, \quad c = \frac{1}{\sqrt{\mu_{0}
\varepsilon_{0}}}.
\]
The same equation holds for $\vect{B}$. Each Cartesian component of
the field obeys the scalar wave equation. The wave speed in vacuum is
$c = 2.997\,924\,58\times 10^{8}\,\si{\meter\per\second}$, fixed
exactly in the 2019 SI redefinition. In a medium with permittivity
$\varepsilon = \varepsilon_{r}\varepsilon_{0}$ and permeability $\mu =
\mu_{r}\mu_{0}$, the wave speed is $v = 1/\sqrt{\mu \varepsilon} =
c/n$, with the \engterm{refractive index} $n = \sqrt{\mu_{r}
\varepsilon_{r}}$. Most non-magnetic media have $\mu_{r}\approx 1$, so
$n\approx \sqrt{\varepsilon_{r}}$; for glass $\varepsilon_{r}\approx
2.3$ at optical frequencies and $n\approx 1.5$, the standard
optical-glass index.

The wave equation is the same equation that governs sound in a fluid
(Volume IV, Chapter 5 acoustics), strings under tension (Volume III,
Chapter 11), and the small-amplitude surface waves of fluids (Volume
III, Chapter 13). The mathematics is shared; the physical content is
domain-specific. Engineers who internalise the wave equation in one
domain transport their intuition to others at low cost.

\section{Plane waves and polarisation}\pathtag{core}

A plane wave is a solution of the wave equation in which the fields
depend only on $\vect{k}\cdot \vect{r} - \omega t$, with $\vect{k}$
the wavevector and $\omega$ the angular frequency. The simplest
sinusoidal plane wave is
\[
\vect{E}(\vect{r}, t) = \vect{E}_{0}\cos(\vect{k}\cdot \vect{r} -
\omega t),
\]
with $|\vect{k}| = \omega/v$ and $\vect{E}_{0}$ a constant
(amplitude) vector. Substituting into Faraday's law gives
\[
\vect{B} = \frac{1}{\omega}\vect{k}\times \vect{E},
\]
so $\vect{B}$ is perpendicular to both $\vect{k}$ and $\vect{E}$, and
$|\vect{B}| = |\vect{E}|/v$. The triplet $(\vect{E}, \vect{B},
\vect{k})$ is right-handed; the wave is transverse (the fields are
perpendicular to the propagation direction).

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-plane-wave-eb.tex}

The \engterm{polarisation} of a plane wave is the orientation of
$\vect{E}_{0}$ in the plane perpendicular to $\vect{k}$. Two
independent polarisations span this plane:
\begin{itemize}[leftmargin=*]
  \item \engterm{Linear polarisation}: $\vect{E}_{0}$ is fixed in
    direction. Light from a thermal source is unpolarised (the sum
    of many uncorrelated emissions of all polarisations); light from
    a laser is highly linearly polarised by the laser's cavity
    geometry.
  \item \engterm{Circular polarisation}: $\vect{E}_{0}$ rotates around
    $\vect{k}$ at angular frequency $\omega$; right-circular if the
    rotation is right-handed about $\vect{k}$, left-circular
    otherwise. Sum of two equal-amplitude linear polarisations
    $90^{\circ}$ apart in space and time, the standard production by
    a quarter-wave plate.
  \item \engterm{Elliptical polarisation}: the general case, in which
    the tip of $\vect{E}$ traces an ellipse in the plane perpendicular
    to $\vect{k}$.
\end{itemize}

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-polarization-ellipse.tex}
A linearly polarised wave passes through a polariser oriented at angle
$\theta$ to the wave's polarisation direction with amplitude
attenuation $\cos\theta$ and intensity attenuation $\cos^{2}\theta$,
the \engterm{Malus law}, derived by Etienne-Louis Malus in 1809. The
$\cos^{2}\theta$ dependence is the working content of polarised-light
filters, LCD displays, and circular-polariser sunglasses.

\subsection{Energy flow and the factor of one half}

The energy a plane wave carries follows from the Poynting vector
$\vect{S} = \vect{E}\times \vect{H}$ introduced in Chapter 8, the
power flux density in $\si{\watt\per\meter\squared}$. For the
travelling wave above, $\vect{H} = \vect{B}/\mu_0$ lies along
$\unitvec{k}\times \vect{E}/(\mu_0 c) = \vect{E}\times \unitvec{k}$
turned through a right angle, with magnitude $H = E/\eta$ and the wave
impedance $\eta = \mu_0 c = \sqrt{\mu_0/\varepsilon_0}$. Because
$\vect{E}$ and $\vect{H}$ are mutually perpendicular and both
transverse, the cross product points along $\unitvec{k}$ and has
magnitude $EH$:
\[
\vect{S}(\vect{r},t) = E(\vect{r},t)\,H(\vect{r},t)\,\unitvec{k} =
\frac{E_0^{2}}{\eta}\cos^{2}(\vect{k}\cdot \vect{r} - \omega t)\,
\unitvec{k}.
\]
The instantaneous flux is the product of two in-phase sinusoids, so it
never reverses sign: a travelling wave carries energy steadily forward,
in pulses at twice the field frequency, as figure
\ref{fig:vol04ch10:poynting-energy} shows. The time average of
$\cos^{2}$ over a cycle is $\tfrac{1}{2}$, which is the entire origin of
the factor of one half that recurs through every intensity, link
budget, and radiation-pressure formula in the chapter:
\[
\langle \vect{S}\rangle = \frac{1}{2}\frac{E_0^{2}}{\eta}\,\unitvec{k}
= \frac{1}{2}\frac{E_0^{2}}{\mu_0 c}\,\unitvec{k}.
\]

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-poynting-energy.tex}

The same picture fixes the energy density. The electric energy density
is $u_E = \tfrac{1}{2}\varepsilon_0 E^{2}$ and the magnetic energy
density is $u_B = \tfrac{1}{2}B^{2}/\mu_0$; substituting $B = E/c$ and
$c^{2} = 1/(\mu_0\varepsilon_0)$ gives $u_B = \tfrac{1}{2}\varepsilon_0
E^{2} = u_E$, so the wave's energy is split equally between the electric
and magnetic fields at every instant. The total instantaneous density
is $u = u_E + u_B = \varepsilon_0 E^{2}$, and its time average is
$\langle u\rangle = \tfrac{1}{2}\varepsilon_0 E_0^{2}$. Dividing the
average flux by the average density returns the speed at which the
energy moves,
\[
\frac{\langle S\rangle}{\langle u\rangle} = \frac{E_0^{2}/(2\eta)}
{\varepsilon_0 E_0^{2}/2} = \frac{1}{\eta\varepsilon_0} = c,
\]
the consistency check that energy in a plane wave travels at the wave
speed, as it must. The momentum density of the field is $\langle
u\rangle/c$, so a wave fully absorbed by a surface delivers a radiation
pressure $\langle S\rangle/c$ and a wave fully reflected delivers twice
that, the relation that sizes a solar sail. The routine
\texttt{code/plane\_wave.py} returns the magnetic amplitude, the wave
impedance $\eta = \sqrt{\mu/\varepsilon}$ (equal to $376.7\,\si{\ohm}$
in vacuum), the intensity, the energy density, and the radiation
pressure for any field amplitude and medium; the worked example below
traces the same arithmetic by hand.

\section{Reflection and refraction}\pathtag{core}

A plane wave incident on a planar boundary between two media of
refractive indices $n_{1}$ and $n_{2}$ produces a reflected wave and a
transmitted (refracted) wave. The boundary conditions of Chapter 8
($\vect{E}$ tangential continuous, $\vect{B}$ normal continuous, with
appropriate jumps in $\vect{D}$ and $\vect{H}$) determine the
amplitudes of the reflected and transmitted waves and the angles at
which they propagate.

The geometric content is \engterm{Snell's law}:
\[
n_{1}\sin\theta_{1} = n_{2}\sin\theta_{2},
\]
where $\theta_{1}$ is the angle of incidence (measured from the
normal to the boundary) and $\theta_{2}$ is the angle of refraction.
Snell's law follows from the requirement that the phase of the field
match across the boundary at every point and at every time; the
algebra is in any standard optics text \cite{text:hecht-optics}, but
the physical content is phase-matching.

For light going from a denser medium to a less dense medium ($n_{1}>
n_{2}$), Snell's law has no solution for $\theta_{2}$ when $\sin
\theta_{1} > n_{2}/n_{1}$; the transmitted wave becomes evanescent
and all the incident power reflects. This is \engterm{total internal
reflection}, the operating principle of optical fibres (Chapter 11)
and of the prism in classical binoculars. The critical angle is
$\theta_{c} = \arcsin(n_{2}/n_{1})$.

The amplitude content is the \engterm{Fresnel equations}. For a wave
with electric field perpendicular to the plane of incidence (s-
polarisation), the reflected amplitude ratio is
\[
r_{s} = \frac{n_{1}\cos\theta_{1} - n_{2}\cos\theta_{2}}{n_{1}\cos
\theta_{1} + n_{2}\cos\theta_{2}}.
\]
For a wave with electric field parallel to the plane of incidence (p-
polarisation),
\[
r_{p} = \frac{n_{2}\cos\theta_{1} - n_{1}\cos\theta_{2}}{n_{2}\cos
\theta_{1} + n_{1}\cos\theta_{2}}.
\]
The intensity reflectivity is $R_{s,p} = |r_{s,p}|^{2}$; at normal
incidence both reduce to
\[
R = \left|\frac{n_{1}-n_{2}}{n_{1}+n_{2}}\right|^{2}.
\]
For a glass-air interface ($n_{1}=1$, $n_{2}=1.5$) at normal
incidence, $R = 0.04$, the $4\,\%$ per-surface loss that drives the
anti-reflection-coating discipline of modern optical systems.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-reflection-boundary.tex}

The Fresnel equations have a Brewster's-angle zero: at $\theta_{B} =
\arctan(n_{2}/n_{1})$, $r_{p} = 0$ and the reflected wave is purely
s-polarised. For a glass-air interface, $\theta_{B} \approx 57^{\circ}$,
the working incidence angle for polarised-light filters and laser-
cavity windows.

\subsection{The Fresnel coefficients from the boundary conditions}

The Fresnel ratios are not postulated; they are the unique amplitudes
that let the fields meet the boundary conditions. We derive the
s-polarised case, where the electric field points out of the plane of
incidence and is therefore tangential to the interface throughout. Write
the incident, reflected, and transmitted electric-field amplitudes as
$E_i$, $E_r$, $E_t$, all along the same tangential direction. The first
condition is continuity of the tangential electric field across the
boundary,
\[
E_i + E_r = E_t.
\]
The second is continuity of the tangential magnetic field. For each plane
wave the magnetic amplitude is $H = E/\eta = nE/\eta_0$ with $\eta_0$ the
free-space impedance, and only the component of $\vect{H}$ along the
surface counts, which introduces the obliquity factor $\cos\theta$ for a
field lying in the plane of incidence. The incident and transmitted waves
carry their tangential $H$ in one sense and the reflected wave in the
opposite sense, so
\[
(E_i - E_r)\,n_1\cos\theta_1 = E_t\,n_2\cos\theta_2,
\]
where Snell's law has already fixed $\theta_2$. Two equations in the two
unknown ratios $r_s = E_r/E_i$ and $t_s = E_t/E_i$ solve at once.
Substituting $E_t = E_i + E_r$ into the magnetic condition and dividing
through by $E_i$,
\[
(1 - r_s)\,n_1\cos\theta_1 = (1 + r_s)\,n_2\cos\theta_2,
\]
and collecting the $r_s$ terms gives
\[
r_s = \frac{n_1\cos\theta_1 - n_2\cos\theta_2}{n_1\cos\theta_1 +
n_2\cos\theta_2}, \quad t_s = 1 + r_s = \frac{2 n_1\cos\theta_1}
{n_1\cos\theta_1 + n_2\cos\theta_2}.
\]
The p-polarised case repeats the construction with the magnetic field
out of the plane of incidence and the electric field carrying the
obliquity factor, which swaps the roles of $n_1$ and $n_2$ in the
numerator and yields the $r_p$ quoted above. The whole content of the
derivation is two continuity statements and one substitution; everything
that follows, including Brewster's angle and total internal reflection,
is a special value of these two ratios.

A check that the algebra is sound is energy conservation. The reflected
and transmitted intensities must sum to the incident intensity, but the
transmitted beam crosses a different area of the interface (it refracts
to a new angle) and travels in a medium of different impedance, so the
transmittance is not simply $|t_s|^2$. Accounting for both effects, the
power transmittance for s-polarisation is
\[
T_s = \frac{n_2\cos\theta_2}{n_1\cos\theta_1}\,|t_s|^2,
\]
and a short manipulation confirms $R_s + T_s = |r_s|^2 + T_s = 1$ at
every angle below the critical angle. The obliquity ratio $n_2\cos
\theta_2/(n_1\cos\theta_1)$ is the geometric correction a beginner most
often drops, turning an apparent energy gain or loss at the interface
into a balanced ledger. Above the critical angle $\cos\theta_2$ turns
imaginary, $|r_s| = 1$, $T_s = 0$, and all the power reflects: the same
two boundary conditions that split the beam at a shallow angle send it
entirely back at a steep one.

\section{Transmission lines}\pathtag{standard}

A \engterm{transmission line} is a two-conductor structure designed to
guide an electromagnetic wave from a source to a load with controlled
impedance and minimal radiation. The canonical examples are coaxial
cable, parallel-plate waveguide (in microwave-board form, a
\engterm{stripline}), parallel-wire (twin-lead, twisted pair), and
microstrip on a PCB. In each, a TEM (transverse electromagnetic) mode
exists in which $\vect{E}$ and $\vect{B}$ are both transverse to the
propagation direction, with no axial field component, propagating at
the medium's speed of light $v = c/\sqrt{\varepsilon_{r}\mu_{r}}$.

The standard model of a transmission line is the
\engterm{distributed-element} circuit: per-unit-length series
inductance $L'$, series resistance $R'$ (loss in the conductors),
shunt capacitance $C'$ (between conductors), shunt conductance $G'$
(loss in the dielectric). The voltage and current along the line obey
the \engterm{telegrapher's equations},
\[
\pdv{V}{z} = -R' I - L'\pdv{I}{t}, \quad \pdv{I}{z} = -G' V - C'
\pdv{V}{t},
\]
which combine to
\[
\frac{\partial^{2} V}{\partial z^{2}} = L'C'\frac{\partial^{2} V}{
\partial t^{2}} + (R'C'+L'G')\pdv{V}{t} + R'G' V,
\]
a damped wave equation in $V(z,t)$. For a lossless line ($R' = G' =
0$), the equation reduces to the standard wave equation with phase
velocity $v = 1/\sqrt{L'C'}$ and characteristic impedance
\[
Z_{0} = \sqrt{L'/C'}.
\]
A coaxial RG-58 cable has $Z_{0} = 50\,\si{\ohm}$; RG-6 video cable
has $Z_{0} = 75\,\si{\ohm}$; the parallel-wire twin lead for
historical television antennas has $Z_{0} = 300\,\si{\ohm}$. The
dataset \texttt{data/coax\_cables.csv} lists characteristic impedance,
velocity factor, and frequency-dependent loss for the common cable
families used in the exercises. The
characteristic impedance is the impedance the source sees if the line
is terminated in a matched load; under that condition no reflection
occurs and the line is in its quietest operating state.

In sinusoidal steady state, the line's input impedance with a load
$Z_{L}$ at the far end of a line of length $\ell$ is
\[
Z_{\text{in}} = Z_{0}\frac{Z_{L} + jZ_{0}\tan(\beta \ell)}{Z_{0} +
jZ_{L}\tan(\beta \ell)}, \quad \beta = \omega/v,
\]
where $\beta$ is the phase constant. Two special cases give the
operating-regime intuition:
\begin{itemize}[leftmargin=*]
  \item Quarter-wave line ($\beta \ell = \pi/2$): $Z_{\text{in}} =
    Z_{0}^{2}/Z_{L}$. A quarter-wave line transforms a short into an
    open and vice versa; with $Z_{0} = \sqrt{Z_{L}Z_{s}}$ it matches
    a load impedance $Z_{L}$ to a source impedance $Z_{s}$, the
    \engterm{quarter-wave transformer}.
  \item Half-wave line ($\beta \ell = \pi$): $Z_{\text{in}} = Z_{L}$.
    A half-wave line repeats the load impedance unchanged.
\end{itemize}

The \engterm{reflection coefficient} at the load is
\[
\Gamma = \frac{Z_{L}-Z_{0}}{Z_{L}+Z_{0}}.
\]
For $Z_{L} = Z_{0}$ (matched load), $\Gamma = 0$ and no reflection
occurs. For $Z_{L} = 0$ (short), $\Gamma = -1$. For $Z_{L} = \infty$
(open), $\Gamma = +1$. The reflection coefficient ranges in magnitude
from $0$ (matched) to $1$ (fully reflective), and its phase ranges
over $360^{\circ}$.

The \engterm{voltage standing wave ratio} (VSWR or SWR) on the line is
\[
\text{VSWR} = \frac{1+|\Gamma|}{1-|\Gamma|},
\]
ranging from $1$ (perfect match) to $\infty$ (open or short). A VSWR
of $2$ corresponds to $|\Gamma| = 1/3$ and a reflected power fraction
of $1/9 \approx 11\,\%$; a VSWR of $1.5$ corresponds to $|\Gamma| =
0.2$ and reflected power of $4\,\%$. Working RF engineers aim for
$\text{VSWR}\le 1.5$ across the operating band; $\text{VSWR} \le 2$ is
acceptable for many applications but begins to stress final-stage
amplifiers and to reveal in the receiver's noise floor.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-standing-wave.tex}

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-mismatch-loss.tex}

The routine \texttt{code/transmission\_line.py} computes the reflection
coefficient, VSWR, reflected and delivered power fractions, return loss,
and the input impedance of a lossless line of any electrical length,
with complex load impedances handled directly.

The \engterm{Smith chart} is the graphical companion to this algebra.
It maps the complex reflection-coefficient plane (the unit disk
$|\Gamma|\le 1$) onto a grid of constant-resistance and constant-
reactance circles, so that a normalised load impedance $z_{L} =
Z_{L}/Z_{0}$ reads off as a single point. Moving along the line toward
the generator rotates that point clockwise around the chart centre at
constant radius, since $|\Gamma|$ is conserved on a lossless line and
only its phase advances by $2\beta z$. A full wavelength of travel is
one full rotation; a quarter wavelength is a half rotation, which sends
a point to its image through the centre and realises the impedance
inversion $Z_{\text{in}} = Z_{0}^{2}/Z_{L}$ geometrically. The
constant-radius circle a point sits on is a constant-VSWR circle, so the
chart reads match quality at a glance. Single-stub and double-stub
matching, the quarter-wave transformer, and lumped-element matching
networks all reduce to constructions on the chart; the simulation
exercise builds a working Smith-chart matcher in code.

\begin{archetype}[Transport across an interface]
A transmission line is the transport archetype made explicit: power
flows from source to load, and the line's characteristic impedance is
the working obstacle to mismatched power transfer. The interface
archetype reappears at every connector, every solder joint, and every
junction between conductor sizes. The discipline of RF design is the
discipline of matched interfaces along an entire transport chain.
\end{archetype}

\section{Waveguides}\pathtag{standard}

A \engterm{waveguide} is a single-conductor structure that confines
the field by reflection from its conducting walls. The canonical
example is the rectangular metal pipe of dimensions $a\times b$ with
$a>b$; the canonical operating mode is the $\text{TE}_{10}$ mode, in
which the electric field is transverse to the propagation axis and
has a single half-wavelength variation across the wider dimension $a$.

\subsection{Modes and cutoff from the wave equation}

A hollow conductor cannot carry a TEM wave, because a TEM field needs
two conductors to hold the transverse voltage between them; a single
pipe supports only modes with an axial field component. We split these
into \engterm{transverse-electric} (TE) modes, with $E_z = 0$ and a
non-zero $H_z$, and \engterm{transverse-magnetic} (TM) modes, with
$H_z = 0$ and a non-zero $E_z$. The whole mode structure follows from
solving the Helmholtz equation for the one axial component and deriving
the rest from it. Seek a field travelling as $e^{-j\beta z}$ along the
guide, so every component carries the factor $e^{j(\omega t - \beta z)}$.
The axial field then obeys the two-dimensional Helmholtz equation in the
cross-section,
\[
\left(\frac{\partial^{2}}{\partial x^{2}} + \frac{\partial^{2}}
{\partial y^{2}} + k_c^2\right)\psi(x,y) = 0, \quad
k_c^2 = \frac{\omega^2}{c^2} - \beta^2,
\]
where $\psi$ stands for $H_z$ (TE) or $E_z$ (TM) and $k_c$ is the
\engterm{cutoff wavenumber}, the transverse part of the total
wavenumber. Separation of variables, $\psi = X(x)Y(y)$, splits this into
two harmonic equations whose solutions are sines and cosines, with the
wall boundary conditions fixing which. For a TE mode the condition is
that the normal derivative of $H_z$ vanish on the walls (so that the
tangential electric field, built from the transverse gradient of $H_z$,
vanishes there), which selects cosines and quantises the transverse
wavenumbers,
\[
H_z(x,y) = H_0\cos\!\left(\frac{m\pi x}{a}\right)
\cos\!\left(\frac{n\pi y}{b}\right), \quad
k_c = \sqrt{\left(\frac{m\pi}{a}\right)^2 +
\left(\frac{n\pi}{b}\right)^2},
\]
with $m,n$ non-negative integers not both zero. For a TM mode the
tangential electric field includes $E_z$ itself, which must vanish on
every wall, selecting sines and forbidding either index from being zero,
so the lowest TM mode is $\text{TM}_{11}$. Each pair $(m,n)$ is a
distinct mode with its own field shape and its own cutoff.

The propagation constant follows from the cutoff wavenumber,
\[
\beta = \sqrt{\frac{\omega^2}{c^2} - k_c^2}.
\]
A real $\beta$ requires $\omega/c > k_c$; below that the radicand turns
negative, $\beta$ becomes imaginary, and the mode decays as $e^{-|\beta|
z}$ without carrying power down the guide. The boundary between
propagation and decay is the \engterm{cutoff frequency} of the mode,
where $\beta = 0$ and the whole wavenumber goes into the transverse
pattern,
\[
f_{c,mn} = \frac{c\,k_c}{2\pi} = \frac{c}{2}\sqrt{\left(\frac{m}{a}
\right)^2 + \left(\frac{n}{b}\right)^2}.
\]
The lowest cutoff belongs to the mode with the smallest $k_c$, which for
$a>b$ is $\text{TE}_{10}$ ($m=1$, $n=0$), giving $f_c = c/(2a)$: the
guide passes nothing until the free-space half wavelength shrinks below
the wide dimension. The next cutoffs, for a guide with $a/b = 2.25$ like
the WR-90 standard, are $\text{TE}_{20}$ at $2f_c$, then $\text{TE}_{01}$,
then the degenerate pair $\text{TE}_{11}$ and $\text{TM}_{11}$, as figure
\ref{fig:vol04ch10:waveguide-mode-chart} lays out. The band between the
$\text{TE}_{10}$ cutoff and the $\text{TE}_{20}$ cutoff is the
\engterm{single-mode region}, where only one mode propagates and the
guide carries a clean, predictable field; running below it kills the
signal and running above it lets a second mode propagate at its own
velocity, smearing a pulse and confounding any phase-sensitive component.
The $2{:}1$ aspect ratio of standard guides is chosen so that this
single-mode region is exactly one octave wide, the largest clean band a
rectangular guide allows.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-waveguide-mode-chart.tex}

A waveguide has a \engterm{cutoff frequency}: below it, no propagating
mode exists, and the wave is evanescent (decays exponentially along
the guide). The cutoff for the $\text{TE}_{10}$ mode is
\[
f_{c} = \frac{c}{2a},
\]
in vacuum. For a WR-90 X-band waveguide ($a = 22.86\,\si{\milli
\meter}$, $b = 10.16\,\si{\milli\meter}$), $f_{c} = 6.56\,\si{\giga
\hertz}$, and the recommended operating band is $8.2$ to $12.4\,\si{
\giga\hertz}$, comfortably above cutoff and below the cutoff of the
next mode $\text{TE}_{20}$ at $13.1\,\si{\giga\hertz}$. Working RF
engineers carry the WR-XX nomenclature and the standard waveguide
bands (S, C, X, Ku, Ka, V, W) for radar, satellite, and millimetre-
wave applications. The dataset \texttt{data/waveguide\_bands.csv}
tabulates the inside dimensions, $\text{TE}_{10}$ cutoff, and
recommended operating band for the common WR designations.

Above cutoff, the propagating wave has a phase velocity larger than
$c$ and a group velocity smaller than $c$:
\[
v_{p} = \frac{c}{\sqrt{1-(f_{c}/f)^{2}}}, \quad v_{g} = c\sqrt{1-
(f_{c}/f)^{2}}.
\]
Energy travels at $v_{g}$; signals propagate at $v_{g}$. The phase
velocity exceeds $c$ but carries no information; the apparent
violation of special relativity is illusory, exactly as it is for
the phase velocity of any dispersive wave.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-waveguide-dispersion.tex}

\subsection{Phase velocity, group velocity, and dispersion}

The two velocities deserve a derivation, because the distinction is the
source of more confusion than any other point in wave engineering. The
\engterm{phase velocity} is the speed at which a single sinusoid's
crests move, $v_p = \omega/k$, read straight off the requirement that
the phase $\omega t - kz$ stay constant. A pure sinusoid carries no
information, since it has run forever and will run forever; to send a
signal we must modulate it, and a modulation is a superposition of
several wavenumbers. The simplest such packet is the sum of two tones of
equal amplitude at wavenumbers $k\pm \Delta k/2$ and the corresponding
frequencies $\omega\pm \Delta\omega/2$. Adding them and using the
product rule for cosines gives
\[
E = \cos\!\left[(k-\tfrac{\Delta k}{2})z - (\omega -
\tfrac{\Delta\omega}{2})t\right] + \cos\!\left[(k+\tfrac{\Delta k}{2})z -
(\omega + \tfrac{\Delta\omega}{2})t\right] = 2\cos\!\left(\frac{\Delta k}
{2}z - \frac{\Delta\omega}{2}t\right)\cos(kz - \omega t).
\]
The result is a fast carrier $\cos(kz-\omega t)$, moving at the phase
velocity $\omega/k$, multiplied by a slow envelope $\cos(\tfrac{\Delta
k}{2}z - \tfrac{\Delta\omega}{2}t)$, moving at $\Delta\omega/\Delta k$,
the pattern of figure \ref{fig:vol04ch10:wave-packet}. The envelope is
where the energy and the information sit, and in the limit of a narrow
band of wavenumbers its speed is the \engterm{group velocity}
\[
v_g = \dv{\omega}{k}.
\]

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-wave-packet.tex}

A medium in which $\omega/k$ depends on $k$ is \engterm{dispersive}, and
in a dispersive medium the two velocities differ. The waveguide is the
clean example. Its dispersion relation is $\omega^2/c^2 = \beta^2 +
k_c^2$, the same relation that produced the cutoff; differentiating
implicitly, $2\omega\,\dd\omega/c^2 = 2\beta\,\dd\beta$, gives $v_g =
\dd\omega/\dd\beta = c^2\beta/\omega = c\sqrt{1-(f_c/f)^2}$, while $v_p =
\omega/\beta = c/\sqrt{1-(f_c/f)^2}$. Their product is fixed,
\[
v_p\,v_g = c^2,
\]
so the phase velocity rises above $c$ by exactly the factor the group
velocity falls below it, and no relativity is violated because the
superluminal quantity is the motion of a crest of a wave that already
fills the guide, not the arrival of any new feature.

Dispersion has a cost beyond the velocity split: it spreads a pulse.
Different frequency components of a real signal travel at different group
velocities, so a pulse launched compact arrives smeared, its leading
frequencies outrunning its trailing ones. The spread per unit length is
governed by the curvature $\dd^2\beta/\dd\omega^2$ of the dispersion
relation, the \engterm{group-velocity dispersion}, which vanishes for a
non-dispersive line and grows near the waveguide cutoff where the curve
bends hardest. A digital link operated close to cutoff therefore limits
its own bit rate through pulse spreading, the same mechanism that limits
an optical fibre's bandwidth-distance product in Chapter 11, which is the
reason a guide is run in the gentle middle of its single-mode band and
not at either edge.

Compared to coaxial cable, hollow rectangular waveguide has lower
loss at microwave frequencies because the field is not confined by a
dielectric and the wall losses scale more slowly with frequency. The
mechanical price is bulk, weight, and a high-frequency cutoff; the
working compromise in modern microwave electronics is to use coaxial
cable up to $20\,\si{\giga\hertz}$, waveguide above. Optical fibre
(Chapter 11) is the dielectric-waveguide cousin of the metal
waveguide, with its own cutoff structure and its own loss-versus-
bandwidth trade-offs.

\section{Antennas}\pathtag{core}

An \engterm{antenna} is a transducer between a guided wave on a
transmission line and a free-space wave (or vice versa). Three
canonical antennas appear in this chapter; the design of more
sophisticated antennas, including arrays, horns, parabolic reflectors,
microstrip patches, and phased arrays, is the working content of
antenna engineering texts \cite{text:balanis-antennas}.

The \engterm{half-wavelength dipole} is a straight conductor of total
length $\lambda/2$, fed at its centre. Its radiation pattern is a
torus around the dipole's axis, with peak directivity $1.64$
(corresponding to $2.15\,\si{\decibel\textsubscript{i}}$) and a
half-power beamwidth in the $E$-plane of $78^{\circ}$. The radiation
resistance is approximately $73\,\si{\ohm}$, which is the reason
$50\,\si{\ohm}$ and $75\,\si{\ohm}$ transmission lines pair acceptably
with dipoles; a perfect match to $73\,\si{\ohm}$ would require a
custom-impedance feedline, which is impractical.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-dipole-pattern.tex}

\subsection{Where the $73\,\si{\ohm}$ comes from}

The radiation resistance is the resistance that, carrying the antenna's
feed-point current, would dissipate the power the antenna actually
radiates. We can recover the canonical $73\,\si{\ohm}$ for the half-wave
dipole from the field pattern alone. A centre-fed dipole carries a
standing current that vanishes at the ends and, for the half-wave case,
peaks at the feed; the resulting far field has the angular dependence
\[
F(\theta) = \frac{\cos\!\left(\tfrac{\pi}{2}\cos\theta\right)}
{\sin\theta},
\]
with $\theta$ measured from the dipole axis. The time-averaged radiated
power density at range $r$ is $\langle S\rangle = \eta_0 |I_0|^{2}
F^{2}(\theta)/(8\pi^{2} r^{2})$, where $I_0$ is the peak (feed-point)
current. Integrating this over a sphere collects the total radiated
power,
\[
P_{\text{rad}} = \oint \langle S\rangle\, r^{2}\,\dd\Omega
= \frac{\eta_0 |I_0|^{2}}{8\pi^{2}} \int_{0}^{2\pi}\!\!\int_{0}^{\pi}
F^{2}(\theta)\,\sin\theta\,\dd\theta\,\dd\phi
= \frac{\eta_0 |I_0|^{2}}{4\pi} \int_{0}^{\pi}
\frac{\cos^{2}\!\left(\tfrac{\pi}{2}\cos\theta\right)}{\sin\theta}\,
\dd\theta.
\]
The remaining integral has no elementary closed form; the substitution
$u = \cos\theta$ turns it into a cosine-integral expression whose value
is $C_{\text{in}}(2\pi)/2 = 1.2188$, where $C_{\text{in}}$ is the entire
cosine integral. With $\eta_0 = 376.7\,\si{\ohm}$,
\[
P_{\text{rad}} = \frac{376.7\,|I_0|^{2}}{4\pi}\,(1.2188)
= 36.5\,|I_0|^{2}.
\]
Setting $P_{\text{rad}} = \tfrac{1}{2}|I_0|^{2} R_r$, the power a
resistance $R_r$ dissipates under the same peak current, gives
\[
R_r = \frac{2 P_{\text{rad}}}{|I_0|^{2}} = \frac{\eta_0}{2\pi}\,
C_{\text{in}}(2\pi) = \frac{376.7}{2\pi}\,(1.2188) = 73.1\,\si{\ohm},
\]
the textbook value, with the matching small reactance $+42.5\,\si{\ohm}$
trimmed to zero by shortening the dipole the few per cent that the
project specifies. The same integral evaluated for an arbitrary length
gives the resistance and reactance of any thin linear antenna, and the
short-dipole limit $\ell\ll\lambda$ reduces it to the $R_r = 20\pi^{2}
(\ell/\lambda)^{2}$ quoted below, since for a short element $F(\theta)
\to \tfrac{1}{2}\sin\theta\,(\pi\ell/\lambda)$ and the integral becomes
elementary.

The routine \texttt{code/dipole\_pattern.py} evaluates the field
pattern $F(\theta) = \cos((\pi/2)\cos\theta)/\sin\theta$, recovers the
peak directivity $1.64$ by numerical integration over the sphere, and
returns the $78^{\circ}$ half-power beamwidth by root-finding, which
matches the analytic values quoted here.

A dipole much shorter than $\lambda/2$ is a \engterm{short dipole};
its radiation resistance is $R_{r} = 20\pi^{2}(\ell/\lambda)^{2}$, in
ohms, with $\ell$ the total length. At $\ell = \lambda/10$, $R_{r}
\approx 2\,\si{\ohm}$; the antenna is hard to match and inefficient.
Short dipoles are the working antenna of medium-wave (AM broadcast)
radio receivers, where the receiver's bandwidth is narrow and the
matching is achieved with a parallel-tuned LC circuit.

A \engterm{loop antenna} is a circular or rectangular conductor whose
diameter is small compared to a wavelength (a small loop) or
comparable to a wavelength (a large loop). The small loop has
radiation resistance
\[
R_{r} = 20\pi^{2}(C/\lambda)^{4},
\]
where $C$ is the loop circumference, and a radiation pattern with
maxima in the plane of the loop. The small loop is a magnetic
antenna; its impedance is dominated by inductance, and resonance is
achieved with a parallel-tuned capacitor. Working applications: AM
radio direction-finding, RFID readers, and the antenna for the older
push-button garage-door openers.

An \engterm{antenna array} is a collection of antennas whose outputs
are combined coherently. The radiation pattern of an array is the
product of the pattern of the individual element and the
\engterm{array factor}, which encodes the phasing and positioning of
the elements. For $N$ identical elements spaced $d$ apart along a line
and fed with a progressive phase shift $\alpha$ between neighbours, the
array factor is the geometric sum
\[
\text{AF}(\theta) = \sum_{n=0}^{N-1} e^{jn\psi}, \quad \psi =
\frac{2\pi d}{\lambda}\sin\theta + \alpha,
\]
whose magnitude is $|\sin(N\psi/2)/\sin(\psi/2)|$. The interelement
phase $\psi$ collects the geometric path difference $(2\pi d/\lambda)
\sin\theta$ and the feed phase $\alpha$ in one variable, so steering the
beam is a matter of dialing $\alpha$. A two-element broadside array with
elements spaced $\lambda/2$ apart and fed in phase ($\alpha = 0$) has a
pattern null in the end-fire direction and a peak broadside, with
$3\,\si{\decibel}$ directivity gain over a single element. Adding a
$90^{\circ}$ feed phase squints the peak off broadside, the elementary
case shown in figure \ref{fig:vol04ch10:array-factor-2el}. Phased arrays
of hundreds or thousands of elements (modern radar, 5G mobile-radio base
stations, satellite communications) achieve beam steering by varying the
phase of each element electronically, with no moving parts.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-array-factor-2el.tex}

The steering rule generalises to any number of elements. To point the
main lobe of a uniform linear array at an angle $\theta_{0}$ off
broadside, the interelement phase $\psi$ must vanish in that direction,
which fixes the feed phase at $\alpha = -(2\pi d/\lambda)\sin\theta_{0}$.
For half-wavelength spacing this is $\alpha = -\pi\sin\theta_{0}$, so a
quarter-period phase step of $-\pi/2$ per element points the beam to
$30^{\circ}$. Figure \ref{fig:vol04ch10:beam-steering} shows an
eight-element array swung from broadside to $30^{\circ}$ by this phase
step alone, with the element positions unchanged; the whole content of
electronic beam steering is in that one substitution. The price is that a
beam steered far off broadside widens, since the array's projected length
shrinks as $\cos\theta_{0}$, and that grating lobes appear if the element
spacing exceeds $\lambda/2$ when steered, the reason phased-array spacings
sit at or below a half wavelength.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-beam-steering.tex}

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-gain-pattern.tex}

The dataset \texttt{data/antenna\_gains.csv} collects representative
gains, beamwidths, and uses for the antenna families named in this
chapter, from the isotropic reference through the short dipole and
half-wave dipole to high-gain Yagi, horn, and parabolic-reflector
antennas; the gain and beamwidth scale together, since a narrower beam
concentrates the same radiated power into a smaller solid angle.

The \engterm{Friis transmission equation} gives the received power
from a transmitter to a receiver in free space:
\[
P_{r} = P_{t}\frac{G_{t} G_{r} \lambda^{2}}{(4\pi R)^{2}},
\]
where $P_{t}$ is the transmitted power, $G_{t}$ and $G_{r}$ are the
transmit and receive antenna gains, $\lambda$ is the wavelength, and
$R$ is the range. The inverse-square law in $R$, the wavelength
factor (longer wavelengths give larger received power for the same
physical aperture, but shorter wavelengths allow larger gains for
the same aperture), and the gain factors together set the link
budget of every wireless system from a TV remote to a deep-space
probe.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-friis-link.tex}

The routine \texttt{code/friis\_link.py} carries the link budget in
both forms: the linear Friis equation and the decibel arithmetic that
working engineers prefer, with free-space path loss, effective aperture,
and dBm conversions as separate calls. The satellite estimate below and
the WiFi exercise both fall out of the same routine.

\subsection{Where the Friis equation comes from}

The Friis equation reads like three unrelated factors bolted together,
and a reader who carries it as a formula will misplace the wavelength
sooner or later. The equation falls out of two physical statements, each
worth its own line. The first is spreading. A transmitter radiating
total power $P_t$ into a sphere puts a power density $P_t/(4\pi R^2)$ on
that sphere at range $R$, since the area of the sphere is $4\pi R^2$ and
power is conserved. An antenna with gain $G_t$ concentrates the radiation
toward its boresight by the factor $G_t$ relative to the isotropic
spread, so the power density on the boresight axis is
\[
S = \frac{P_t G_t}{4\pi R^2}.
\]
The gain is defined as exactly this ratio of on-axis density to the
isotropic average, which is why it enters as a clean multiplier. The
second statement is collection. A receive antenna pulls power out of the
passing wave over an \engterm{effective aperture} $A_{\text{eff}}$, an
area that need not match the antenna's physical size, so the received
power is
\[
P_r = S\, A_{\text{eff}} = \frac{P_t G_t A_{\text{eff}}}{4\pi R^2}.
\]
The whole content of the Friis equation is now in the relation between a
receive antenna's effective aperture and its gain. That relation is not a
separate empirical fact; it follows from reciprocity. An antenna's
transmit gain pattern and its receive aperture pattern are the same
function of angle, because Maxwell's equations are symmetric under
interchange of source and field, and a thermodynamic argument fixes the
constant of proportionality. Place any antenna inside a cavity in thermal
equilibrium at temperature $T$. The antenna absorbs the blackbody
radiation that falls on its effective aperture and re-radiates through its
gain pattern, and detailed balance demands that the two rates match in
every direction and at every frequency. Carrying the blackbody spectral
density through the balance yields one relation, independent of the
antenna,
\[
A_{\text{eff}} = \frac{\lambda^2}{4\pi}\,G,
\]
the bridge between the two pictures. The wavelength-squared is the
footprint of a single radiation mode: a diffraction-limited beam fills a
solid angle of order $\lambda^2/A_{\text{eff}}$, and an antenna cannot
couple to a patch of wavefront smaller than that mode. Substituting the
bridge into the collection equation gives the Friis equation in its
standard form,
\[
P_r = \frac{P_t G_t}{4\pi R^2}\cdot \frac{G_r \lambda^2}{4\pi} =
P_t\,\frac{G_t G_r \lambda^2}{(4\pi R)^2}.
\]
The $\lambda^2$ now reads correctly. It is a property of the
gain-to-aperture bridge, not of the empty space the wave crosses, so for
two antennas of fixed physical aperture (where $A_{\text{eff}}$ is set by
the hardware and $G = 4\pi A_{\text{eff}}/\lambda^2$ rises as
$\lambda^{-2}$) the received power actually grows toward higher frequency.
The apparent penalty of short wavelengths in the bare formula is an
artefact of holding gain fixed while the physical aperture shrinks. A
deep-space link holds the dish sizes fixed and prefers the highest
frequency the atmosphere allows; a broadcast link holds the antenna gains
modest and pays the wavelength term. Reading which quantity is held fixed
is the whole skill of applying the equation.

\begin{estimation}
\textbf{Question.} A geostationary satellite transmits a $20\,\si{
\watt}$ signal at $12\,\si{\giga\hertz}$ to a domestic satellite dish
on Earth. The satellite's antenna gain is $35\,\si{\decibel
\textsubscript{i}}$, and the dish has a diameter of $0.6\,\si{\meter}$.
Estimate the received power.

\textbf{Estimate, before reading on.} State a factor-of-three-
confidence answer before reading the working.

\textbf{Working.} The wavelength is $\lambda = c/f = 0.025\,\si{
\meter}$. The geostationary orbit altitude is $35\,786\,\si{\kilo
\meter}$; slant range to a low-latitude receiver is $\approx 38\,000
\,\si{\kilo\meter} = 3.8\times 10^{7}\,\si{\meter}$. The free-space-
path loss term is
\[
L_{\text{fs}} = (4\pi R/\lambda)^{2} = (4\pi\times 3.8\times 10^{7}/
0.025)^{2} \approx 3.6\times 10^{20},
\]
or $206\,\si{\decibel}$. The receive dish has effective aperture area
$A_{\text{eff}} = 0.55\times \pi(0.3)^{2} = 0.16\,\si{\meter}^{2}$
(taking aperture efficiency $0.55$); its gain at $12\,\si{\giga\hertz}$
is $G_{r} = 4\pi A_{\text{eff}}/\lambda^{2} = 4\pi\times 0.16/
(0.025)^{2} \approx 3200$, or $35\,\si{\decibel\textsubscript{i}}$.

The link-budget arithmetic in decibels: $P_{r}\,[\si{\decibel}\text{m}]
= P_{t}\,[\si{\decibel}\text{m}] + G_{t} + G_{r} - L_{\text{fs}} = 43
+ 35 + 35 - 206 = -93\,\si{\decibel}\text{m}$, i.e., $5\times 10^{-13}
\,\si{\watt} = 0.5\,\si{\pico\watt}$.

\textbf{Postmortem.} The received power is much less than the
microwatt the reader might naively guess; the recovery is in the
receive antenna gain and the low-noise amplifier of the dish's LNB.
A reader who guessed within a factor of $100$ on the received power
was already in the working range for satellite-link design; the
inverse-square law over distances of tens of thousands of kilometres
dominates the budget. The numerical example is for direct-broadcast
satellite TV, current as of 2025.
\end{estimation}

\section{Worked examples}\pathtag{standard}

A run of worked examples fixes the daily arithmetic of the chapter in
mind. The first three are the hand computation behind the code routines;
the next three carry the recurring small calculations of polarisation
matching, narrowband impedance matching, and shielding that working RF
engineers do on the back of a notebook. The next three carry the
radar-and-array arithmetic that the second case study leans on: a
Doppler velocity from a frequency shift, the near-field boundary that
sets how far away a measurement must stand, and the guide wavelength
that sizes every waveguide component. The next four reach into the
material side of the chapter: how a wave dies in a conductor and what
that costs an undersea link, when an array spacing breeds a second
unwanted beam, the field strength a phone radiates against an exposure
limit, and the reflection that a polariser turns to its advantage. The
last five close on the core theory the chapter deepened: the group delay
through a near-cutoff guide, the frequency choice that follows from
holding apertures fixed, an energy-balanced Fresnel ledger, the pulse
spreading bred by a second waveguide mode, and the reading of a
slotted-line measurement.

\paragraph{A plane wave at the surface of the Earth.}
Bright sunlight delivers about $1\,\si{\kilo\watt\per\meter\squared}$ at
the ground. Treat it as a single plane wave to estimate its field
amplitudes. The intensity is $I = \tfrac{1}{2}E_{0}^{2}/\eta_{0}$ with
$\eta_{0} = 376.7\,\si{\ohm}$, so
\[
E_{0} = \sqrt{2 I \eta_{0}} = \sqrt{2\times 1000\times 376.7}
\approx 868\,\si{\volt\per\meter},
\]
and the magnetic amplitude is $B_{0} = E_{0}/c \approx 2.9\,\si{\micro
\tesla}$, comparable to the Earth's static field. The radiation pressure
on an absorbing surface is $I/c \approx 3.3\,\si{\micro\pascal}$, the
order of magnitude that solar sails work with. Sunlight is broadband and
unpolarised, so the single-plane-wave picture is a model for the
amplitudes, not a literal description; \texttt{code/plane\_wave.py}
returns each quantity.

\paragraph{Reading a VSWR meter.}
A bench meter reads $\text{VSWR} = 2.5$ on a $50\,\si{\ohm}$ system.
The reflection-coefficient magnitude is $|\Gamma| = (\text{VSWR}-1)/
(\text{VSWR}+1) = 1.5/3.5 = 0.43$, the reflected power fraction is
$|\Gamma|^{2} = 0.18$, so $18\,\%$ of the forward power returns to the
source and $82\,\%$ reaches the load. The return loss is $-20\log_{10}
0.43 = 7.4\,\si{\decibel}$. A $2.5{:}1$ mismatch is past the usual
$1.5{:}1$ design target and into the range where a high-power final
stage needs protection. The same numbers come from
\texttt{code/transmission\_line.py}.

\paragraph{A short microwave hop.}
A point-to-point link at $5.8\,\si{\giga\hertz}$ runs $1\,\si{\kilo
\meter}$ between two $20\,\si{\decibel\textsubscript{i}}$ panel
antennas, with $30\,\si{\decibel}\text{m}$ ($1\,\si{\watt}$) at the
transmitter. The wavelength is $\lambda = 0.0517\,\si{\meter}$, so the
free-space path loss is $20\log_{10}(4\pi\times 1000/0.0517) =
107.7\,\si{\decibel}$. The received power is
\[
P_{r} = 30 + 20 + 20 - 107.7 = -37.7\,\si{\decibel}\text{m},
\]
about $0.17\,\si{\micro\watt}$, comfortably above a typical
$-85\,\si{\decibel}\text{m}$ receiver sensitivity, leaving roughly
$47\,\si{\decibel}$ of link margin for rain, misalignment, and fading.
\texttt{code/friis\_link.py} reproduces the budget.

\paragraph{Polarisation-mismatch loss.}
A circularly polarised transmit antenna illuminates a linearly polarised
receive antenna, the common situation when a satellite radiates circular
polarisation (chosen so that the link survives Faraday rotation in the
ionosphere) and the ground terminal is a plain linear dipole. The
polarisation-mismatch factor is $\rho = |\unitvec{e}_{t}\cdot
\unitvec{e}_{r}^{*}|^{2}$, the squared magnitude of the inner product of
the two unit polarisation vectors. A circular wave decomposes into two
equal orthogonal linear components, so a linear receiver captures exactly
half the power: $\rho = \tfrac{1}{2}$, a flat $3\,\si{\decibel}$ loss
independent of how the dipole is rotated about the line of sight. Two
linear antennas misaligned by an angle $\psi$ instead give $\rho =
\cos^{2}\psi$, the same Malus-law form met in polarised optics; a
$30^{\circ}$ misalignment costs $\cos^{2}30^{\circ} = 0.75$, or
$1.25\,\si{\decibel}$, while a cross-polarised pair ($\psi = 90^{\circ}$)
gives $\rho = 0$ and a nominally infinite loss, in practice limited to
the antennas' finite cross-polarisation rejection of $20$ to
$30\,\si{\decibel}$. The lesson for a link budget is to carry the
polarisation-mismatch term explicitly: the $3\,\si{\decibel}$ circular-to-
linear penalty is a fixed line item, not a margin to be wished away.

\paragraph{A quarter-wave matching section.}
A patch antenna presents a real input impedance of $200\,\si{\ohm}$ at its
resonant frequency of $2.4\,\si{\giga\hertz}$, and we must feed it from a
$50\,\si{\ohm}$ microstrip line. A quarter-wave transformer of
characteristic impedance $Z_{0t} = \sqrt{Z_{s}Z_{L}} = \sqrt{50\times 200}
= 100\,\si{\ohm}$ does the job. At $2.4\,\si{\giga\hertz}$ the free-space
wavelength is $\lambda = c/f = 0.125\,\si{\meter}$; on a microstrip with
effective permittivity $\varepsilon_{\text{eff}} = 6.5$ the guided
wavelength is $\lambda_{g} = \lambda/\sqrt{\varepsilon_{\text{eff}}} =
0.125/2.55 = 0.049\,\si{\meter}$, so the transformer length is
$\lambda_{g}/4 = 12.3\,\si{\milli\meter}$. At the design frequency the
match is perfect, $\Gamma = 0$; one octave is far too wide to ask of a
single section, but across a $\pm 5\,\%$ band the input reflection stays
below $|\Gamma| = 0.1$ (VSWR $1.22$), comfortably inside the usual target.
The same geometric-mean rule, in the field picture of the mastery box
below, is the anti-reflection coating of an optical lens.

\paragraph{Skin-depth shielding thickness.}
An enclosure must attenuate a $1\,\si{\giga\hertz}$ field by
$60\,\si{\decibel}$ through absorption in a copper wall. The skin depth in
copper ($\sigma = 5.96\times 10^{7}\,\si{\siemens\per\meter}$, $\mu =
\mu_{0}$) is $\delta = 1/\sqrt{\pi f \mu \sigma} = 1/\sqrt{\pi(10^{9})
(4\pi\times 10^{-7})(5.96\times 10^{7})} = 2.1\,\si{\micro\meter}$, the
value marked in figure \ref{fig:vol04ch10:skin-depth}. The field decays as
$e^{-x/\delta}$, so the absorption loss in decibels is $A = 20\log_{10}
e^{x/\delta} = 8.69\,(x/\delta)$. Setting $A = 60\,\si{\decibel}$ gives
$x = (60/8.69)\,\delta = 6.9\,\delta = 14.5\,\si{\micro\meter}$. A copper
foil a hundredth of a millimetre thick already provides this absorption at
$1\,\si{\giga\hertz}$; the engineering difficulty of shielding lives not in
the bulk metal but in the seams, gaskets, and cable penetrations, where
leakage through apertures dominates the bulk absorption computed here.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-skin-depth.tex}

\paragraph{A Doppler speed gun.}
A police traffic radar at $24.125\,\si{\giga\hertz}$ (the K band used by
hand-held units) measures a downshifted return from a car receding along
the line of sight, with a beat frequency of $2.30\,\si{\kilo\hertz}$
between the transmitted and received signals. The two-way Doppler shift
for a closing or opening speed $v$ along the beam is $f_{d} = 2 v f/c$,
the factor of two arising because the moving target both receives a
shifted wave and re-radiates a second shift. Solving for the radial
speed,
\[
v = \frac{c\, f_{d}}{2 f} = \frac{(3\times 10^{8})(2.30\times 10^{3})}
{2(24.125\times 10^{9})} = 14.3\,\si{\meter\per\second},
\]
about $51\,\si{\kilo\meter\per\hour}$. The measured speed is the
projection of the true velocity onto the beam, so a gun aimed at a
cosine angle $\phi$ off the line of travel reads $v\cos\phi$ and always
underreads, the geometric reason the cosine error favours the driver
rather than the officer. A $10^{\circ}$ squint costs only
$1.5\,\%$; a $30^{\circ}$ squint costs $13\,\%$, enough to matter in a
dispute. The same $f_{d} = 2 v f/c$ relation, read in the other
direction, sizes the pulse-repetition frequency a pulse-Doppler radar
needs to measure a given closing speed without ambiguity.

\paragraph{The near-field boundary of an aperture.}
A measurement or a link assumes the receiver sits in the far field, where
the wavefront is locally planar and the antenna's gain pattern has
settled into its published shape. For an aperture of largest dimension
$D$ the conventional far-field boundary is the \engterm{Fraunhofer
distance}
\[
R_{\mathrm{ff}} = \frac{2 D^{2}}{\lambda},
\]
the range at which the path-length difference across the aperture falls
below $\lambda/16$. Take the $0.9\,\si{\meter}$ domestic dish of the
first case study at $12\,\si{\giga\hertz}$, where $\lambda = 0.025\,\si{
\meter}$:
\[
R_{\mathrm{ff}} = \frac{2(0.9)^{2}}{0.025} = 64.8\,\si{\meter}.
\]
A satellite tens of thousands of kilometres away is comfortably in the
far field, so the link budget is sound; an antenna range trying to
measure that dish's pattern, though, must place the source at least
$65\,\si{\meter}$ away, or use a compact range with a collimating
reflector, or measure in the near field and transform the result. A
$3\,\si{\meter}$ ground-station dish at the same frequency pushes the
boundary to $720\,\si{\meter}$, which is why large reflectors are
characterised on outdoor ranges or by near-field scanning rather than in
any anechoic chamber that would fit in a building.

\paragraph{A waveguide component sized in guide wavelengths.}
A waveguide bandpass filter or a directional coupler is built from
sections measured in \engterm{guide wavelengths}, which exceed the
free-space wavelength because the wave zig-zags down the guide. In a
WR-90 X-band guide ($a = 22.86\,\si{\milli\meter}$, $f_{c} = 6.56\,\si{
\giga\hertz}$) operated at $10\,\si{\giga\hertz}$, the free-space
wavelength is $\lambda = c/f = 0.030\,\si{\meter}$, and the guide
wavelength is
\[
\lambda_{g} = \frac{\lambda}{\sqrt{1 - (f_{c}/f)^{2}}} = \frac{0.030}
{\sqrt{1 - (6.56/10)^{2}}} = \frac{0.030}{0.752} = 0.0399\,\si{\meter},
\]
about a third longer than free space. A quarter-wave coupling section is
therefore $\lambda_{g}/4 = 9.97\,\si{\milli\meter}$, not the
$7.5\,\si{\milli\meter}$ a free-space estimate would give, and a designer
who used the wrong wavelength would centre the component a band away from
its target. The guide wavelength also sets the longitudinal spacing of
coupling slots in a slotted-waveguide array antenna, where alternate
slots are spaced $\lambda_{g}/2$ apart and tilted to feed in phase.

\paragraph{A wave dying in seawater.}
Seawater is a good conductor at low radio frequencies, with conductivity
$\sigma \approx 4\,\si{\siemens\per\meter}$ and relative permittivity
$\varepsilon_r \approx 81$. A submarine communication signal at $f =
3\,\si{\kilo\hertz}$ (the ELF band) faces the loss-tangent test
$\sigma/(\omega\varepsilon_0\varepsilon_r) = 4/(2\pi(3000)(8.85\times
10^{-12})(81)) = 2.96\times 10^{5}$, far above unity, so the medium is a
good conductor and the wave attenuates over a skin depth $\delta = 1/
\sqrt{\pi f\mu_0\sigma} = 1/\sqrt{\pi(3000)(4\pi\times 10^{-7})(4)} =
4.6\,\si{\meter}$. The field falls by $1/e$ every $4.6\,\si{\meter}$,
or $8.69/\delta = 1.9\,\si{\decibel\per\meter}$, so reaching a boat at
$100\,\si{\meter}$ depth costs about $190\,\si{\decibel}$ of one-way
attenuation. Raising the frequency to $30\,\si{\kilo\hertz}$ (VLF) drops
the skin depth to $1.5\,\si{\meter}$ and the depth penalty triples,
which is the physics behind the choice of kilometre-scale ELF antennas
and a few-bits-per-minute data rate for the deepest submarine links: the
conductor sets the band, the band sets the antenna, and the antenna sets
the throughput.

\paragraph{The onset of a grating lobe.}
A uniform linear array steered to broadside keeps a single main beam as
long as the element spacing $d$ stays at or below $\lambda$, but a
steered array fails sooner. A grating lobe appears when the
interelement phase $\psi = (2\pi d/\lambda)\sin\theta + \alpha$ reaches
$\pm 2\pi$ for some real angle $\theta$, since then a second direction
satisfies the in-phase condition. For a beam steered to $\theta_0$ the
no-grating-lobe condition is $d/\lambda < 1/(1 + |\sin\theta_0|)$. An
array meant to scan to $\theta_0 = 60^{\circ}$ therefore needs $d/
\lambda < 1/(1 + 0.866) = 0.536$, just over a half wavelength, which is
the reason phased-array lattices are built on a half-wavelength pitch:
it is the largest spacing that keeps a single beam across a full
$\pm 60^{\circ}$ scan. Spacing the elements at $0.7\lambda$ to win
aperture would breed a grating lobe at $\sin\theta_g = \sin 60^{\circ} -
\lambda/d = 0.866 - 1.43 = -0.56$, that is a spurious beam at
$-34^{\circ}$ radiating as much power as the wanted one, the classic
self-inflicted blind spot of an over-spread array.

\paragraph{A handset field strength against the exposure limit.}
A mobile handset radiates $P_t = 0.25\,\si{\watt}$ (a typical
power-controlled uplink level) from a near-isotropic internal antenna.
At a bystander distance of $r = 0.2\,\si{\meter}$ the far-field power
density is $S = P_t/(4\pi r^{2}) = 0.25/(4\pi(0.2)^{2}) = 0.50\,\si{
\watt\per\meter\squared}$, and the corresponding field strength follows
from $S = E_{\text{rms}}^{2}/\eta_0$, giving $E_{\text{rms}} =
\sqrt{S\eta_0} = \sqrt{0.50\times 376.7} = 13.7\,\si{\volt\per\meter}$.
The ICNIRP 2020 general-public reference level for power density in the
$2$ to $300\,\si{\giga\hertz}$ range is $10\,\si{\watt\per\meter
\squared}$, so the radiated far-field density at $0.2\,\si{\meter}$ sits
a factor of twenty below the limit. The exposure that governs a handset
is not this radiated far field at all but the near-field absorption in
the head, measured as a specific absorption rate against a separate
$2\,\si{\watt\per\kilogram}$ basic restriction; the far-field estimate
here bounds the bystander case, while the user case demands the SAR
measurement that the failure section describes.

\paragraph{A polarising window at Brewster's angle.}
A laser-cavity window is set at Brewster's angle so that p-polarised
light passes with zero reflection loss, leaving the cavity free to build
a polarised mode. For a fused-silica window of index $n_2 = 1.46$ in air
($n_1 = 1$), Brewster's angle is $\theta_B = \arctan(n_2/n_1) =
\arctan 1.46 = 55.6^{\circ}$, and at that incidence the refraction angle
is $\theta_2 = \arctan(n_1/n_2) = 34.4^{\circ}$, the two summing to
$90^{\circ}$ as Brewster geometry requires. The p-polarised reflectance
there is exactly zero, while the s-polarised reflectance is $R_s =
|r_s|^{2}$ with $r_s = (n_1\cos\theta_B - n_2\cos\theta_2)/(n_1\cos
\theta_B + n_2\cos\theta_2) = (0.564 - 1.205)/(0.564 + 1.205) = -0.362$,
so $R_s = 0.131$. The window passes p-polarisation without loss and
reflects $13\,\%$ of the orthogonal polarisation per surface, which over
many cavity round trips starves the s-mode and lets the p-mode dominate;
the same $13\,\%$ reflected beam, taken off to the side, is a clean
linearly polarised pick-off for a beam monitor.

\paragraph{Group delay through a near-cutoff guide.}
A pulse travels through a $0.30\,\si{\meter}$ section of WR-90 X-band
guide ($f_c = 6.56\,\si{\giga\hertz}$) at two carrier frequencies, one
near cutoff and one well above, and we compare the transit times. The
group velocity is $v_g = c\sqrt{1-(f_c/f)^2}$, so the delay over a length
$L$ is $t = L/v_g$. At $7.0\,\si{\giga\hertz}$, just above cutoff,
$\sqrt{1-(6.56/7.0)^2} = \sqrt{1-0.878} = 0.349$, so $v_g = 1.05\times
10^{8}\,\si{\meter\per\second}$ and the delay is $t = 0.30/(1.05\times
10^{8}) = 2.86\,\si{\nano\second}$. At $11.0\,\si{\giga\hertz}$, near the
middle of the band, $\sqrt{1-(6.56/11.0)^2} = 0.803$, so $v_g = 2.41
\times 10^{8}\,\si{\meter\per\second}$ and the delay is $1.25\,\si{\nano
\second}$. The near-cutoff carrier takes more than twice as long over the
same metal, and the delay climbs without bound as $f\to f_c$, where the
group velocity falls to zero. A component built to a fixed phase length
at one frequency therefore mistimes badly at another near cutoff, the
reason a wideband waveguide system is kept clear of the low band edge and
why the steep delay slope there is read as group-velocity dispersion.

\paragraph{Holding the apertures fixed: a deep-space frequency choice.}
A deep-space link uses a $70\,\si{\meter}$ ground dish and a fixed
$3\,\si{\meter}$ spacecraft dish, and a designer must choose between the
S band ($2.3\,\si{\giga\hertz}$) and the X band ($8.4\,\si{\giga\hertz}$).
The naive Friis reading, with its $\lambda^2$ in the numerator, seems to
favour the lower frequency, but the apertures are fixed hardware, so the
gains rise with frequency as $G = 4\pi A_{\text{eff}}/\lambda^2$. The
received power is $P_r = P_t A_{t,\text{eff}} A_{r,\text{eff}}/(\lambda^2
R^2)$ once both gains are written through their apertures, so for fixed
apertures $P_r\propto 1/\lambda^2 \propto f^2$. Moving from S to X band
multiplies the frequency by $8.4/2.3 = 3.65$, so the received power rises
by $3.65^2 = 13.3$, or $11.2\,\si{\decibel}$, with not one square metre of
new antenna. This is the whole reason planetary missions migrated from S
to X band and now reach toward the Ka band: the higher frequency wins
because the dishes, not the gains, are the fixed cost. The countervailing
penalties are tighter pointing (the beam narrows as $1/f$) and more rain
and atmospheric loss at the ground station, the trade that stops the climb
short of the optical.

\paragraph{An energy-balanced Fresnel ledger.}
Light in air ($n_1 = 1$) strikes glass ($n_2 = 1.5$) at $\theta_1 =
60^{\circ}$, s-polarised, and we verify that the reflected and
transmitted powers sum to the incident. Snell's law gives $\sin\theta_2 =
\sin 60^{\circ}/1.5 = 0.577$, so $\theta_2 = 35.3^{\circ}$, with
$\cos\theta_1 = 0.500$ and $\cos\theta_2 = 0.816$. The s-reflection
coefficient is
\[
r_s = \frac{n_1\cos\theta_1 - n_2\cos\theta_2}{n_1\cos\theta_1 +
n_2\cos\theta_2} = \frac{0.500 - 1.224}{0.500 + 1.224} = \frac{-0.724}
{1.724} = -0.420,
\]
so the reflectance is $R_s = |r_s|^2 = 0.176$. The transmission
coefficient is $t_s = 1 + r_s = 0.580$, and the transmittance carries the
obliquity-and-impedance factor,
\[
T_s = \frac{n_2\cos\theta_2}{n_1\cos\theta_1}\,|t_s|^2 = \frac{1.5\times
0.816}{1\times 0.500}\,(0.580)^2 = 2.448\times 0.336 = 0.824.
\]
The sum is $R_s + T_s = 0.176 + 0.824 = 1.000$, the ledger balancing
exactly. A reader who forgot the obliquity factor and wrote $T_s =
|t_s|^2 = 0.336$ would find $R_s + T_s = 0.512$ and conclude, wrongly,
that the interface destroyed half the power; the geometric factor is the
whole difference between a sound calculation and a nonsensical one.

\paragraph{Pulse spreading from a second waveguide mode.}
A short pulse at $9.0\,\si{\giga\hertz}$ is launched into a $5\,\si{
\meter}$ guide that, through a fabrication error, supports both the
$\text{TE}_{10}$ mode ($f_{c1} = 6.56\,\si{\giga\hertz}$) and the
$\text{TE}_{20}$ mode ($f_{c2} = 13.1\,\si{\giga\hertz}$). Only the first
mode propagates here, since $9.0\,\si{\giga\hertz}$ is below the second
cutoff, so the pulse stays clean, but suppose the same pulse were sent at
$14.0\,\si{\giga\hertz}$, above both cutoffs, where both modes carry
power. The two modes travel at different group velocities, $v_{g1} =
c\sqrt{1-(6.56/14)^2} = 0.883c$ and $v_{g2} = c\sqrt{1-(13.1/14)^2} =
0.353c$. Over $5\,\si{\meter}$ the arrival times are $t_1 = 5/(0.883c) =
18.9\,\si{\nano\second}$ and $t_2 = 5/(0.353c) = 47.2\,\si{\nano\second}$,
a spread of $28.3\,\si{\nano\second}$ between the two copies of the pulse.
A receiver expecting one pulse sees two, separated by nearly $30\,\si{
\nano\second}$, which corrupts any data faster than about $35\,\si{\mega
\bit\per\second}$ on that link. The lesson is operational: a guide run
above its single-mode band carries a mode-dispersion penalty that grows
with length, and the cure is to stay below the $\text{TE}_{20}$ cutoff so
that only one velocity exists.

\paragraph{Reading a slotted-line measurement.}
Before vector network analysers, an RF engineer measured an unknown load
with a \engterm{slotted line}, a length of air guide with a sliding probe
that samples the standing-wave voltage. A measurement at $3\,\si{\giga
\hertz}$ on a $50\,\si{\ohm}$ slotted line reads a voltage maximum of
$4.0\,\si{\milli\volt}$ and an adjacent minimum of $1.0\,\si{\milli
\volt}$, with the minima spaced $50\,\si{\milli\meter}$ apart. The
standing-wave ratio is the ratio of the readings, $\text{VSWR} = V_{\max}/
V_{\min} = 4.0/1.0 = 4.0$, so the reflection magnitude is $|\Gamma| =
(\text{VSWR}-1)/(\text{VSWR}+1) = 3/5 = 0.60$ and the reflected power
fraction is $0.36$. The spacing of the minima is a half guide wavelength,
so $\lambda_g = 2\times 50 = 100\,\si{\milli\meter}$, which for an air
line equals the free-space wavelength $\lambda = c/f = 100\,\si{\milli
\meter}$, a consistency check that the line is air-filled. The position of
a minimum relative to the load, in fractions of $\lambda_g$, fixes the
phase of $\Gamma$ and so completes the load impedance; a minimum at the
load itself would mean a purely real load below $Z_0$, and a maximum at
the load a purely real load above it. The slotted line is obsolete, but
the arithmetic, VSWR from a max-to-min ratio and wavelength from the null
spacing, is exactly what a modern analyser computes internally.

\begin{mastery}
The wave impedance $\eta = \sqrt{\mu/\varepsilon}$ and the
characteristic impedance $Z_{0} = \sqrt{L'/C'}$ are the same idea in
two guises. Both are the ratio of a ``through'' quantity to an
``across'' quantity for a forward-travelling wave: $E/H$ for a plane
wave in a medium, $V/I$ for a wave on a line. The reflection
coefficient at any interface is $(\eta_{2}-\eta_{1})/(\eta_{2}+
\eta_{1})$ in the field picture and $(Z_{L}-Z_{0})/(Z_{L}+Z_{0})$ in
the circuit picture, the identical algebraic form. This is why the
optical anti-reflection coating and the quarter-wave impedance
transformer are one device: a quarter-wave layer of intermediate
impedance $\sqrt{\eta_{1}\eta_{2}}$ cancels the reflection at a single
frequency, whether the wave is light in glass or a signal on a line.
The unification is the working reason an RF engineer and an optical
engineer can read each other's matching problems.
\end{mastery}

\begin{mastery}
The Friis equation has a cousin that governs radar, and it follows from
the same effective-aperture idea together with the reciprocity of
antennas. \engterm{Reciprocity} states that an antenna's transmit gain
pattern and its receive pattern are identical, a consequence of the
symmetry of Maxwell's equations under source-field interchange; the
working content is the relation between gain and effective aperture,
$A_{\text{eff}} = G\lambda^{2}/(4\pi)$, which holds whether the antenna
sends or receives. The \engterm{radar range equation} chains two
Friis-style hops. A radar of transmit power $P_{t}$ and antenna gain
$G$ illuminates a target at range $R$ with power density $P_{t}G/(4\pi
R^{2})$. The target intercepts and re-radiates this with a
\engterm{radar cross-section} $\sigma$ (units of area), scattering an
effective power $P_{t}G\sigma/(4\pi R^{2})$ that spreads back over a
second inverse-square hop. The same antenna, with effective aperture
$A_{\text{eff}} = G\lambda^{2}/(4\pi)$, collects the echo, so the
received power is
\[
P_{r} = \frac{P_{t}\,G^{2}\,\lambda^{2}\,\sigma}{(4\pi)^{3}\,R^{4}}.
\]
The defining feature is the $R^{-4}$ dependence: doubling the detection
range demands sixteen times the transmit power, or four times the
antenna gain, which is why long-range radars carry both megawatt
transmitters and very large apertures. Setting $P_{r}$ to the receiver's
minimum detectable signal and solving for $R$ gives the radar's maximum
range, the single number a radar engineer designs toward. The same
$R^{-4}$ law sets the read range of a passive RFID tag, which is a
backscatter radar in miniature.
\end{mastery}

\begin{mastery}
The radar range equation in its bare form hides where the detection
sensitivity comes from. A single pulse arrives buried in receiver noise
of power $N = k T_{\mathrm{sys}} B$, where $B$ is the receiver bandwidth,
so the single-pulse signal-to-noise ratio is the echo power divided by
$N$. Two devices buy back sensitivity that the $R^{-4}$ spreading takes
away. The first is the \engterm{matched filter}: a receiver whose
frequency response is the complex conjugate of the transmitted pulse
spectrum maximises the output signal-to-noise ratio, and the maximum it
delivers is $2E/N_{0}$, the pulse energy $E = P_{t}\tau$ over the noise
density, independent of the pulse shape. A long pulse therefore detects a
weak target as well as a short high-power pulse of the same energy, which
is why radars transmit coded pulses and compress them on receive, trading
peak power for average power. The second is \engterm{integration}:
summing $n$ successive echoes coherently raises the signal-to-noise ratio
by a factor $n$ (by $10\log_{10} n$ in decibels), since the signal adds
in amplitude while the noise adds in power; integrating non-coherently,
after the detector, gives a smaller gain near $\sqrt{n}$. A radar that
dwells on a target for $32$ pulses gains $15\,\si{\decibel}$ coherently,
which by the $R^{-4}$ law extends the detection range by a factor of
$10^{15/40} = 2.4$. The detection threshold itself is set by the
tolerated false-alarm probability through the noise statistics, the
content of the detection curve in figure
\ref{fig:vol04ch10:detection-curve}: demanding a rarer false alarm raises
the threshold and costs SNR, the trade a radar designer balances against
the cost of missing a real target.
\end{mastery}

\begin{mastery}
A small antenna cannot be made arbitrarily wideband, and the reason is a
fundamental limit rather than a craft shortfall. An antenna that fits
inside a sphere of radius $a$ stores reactive energy in the near field
that does not radiate; the ratio of stored energy to radiated power per
cycle is a quality factor $Q$, and a high $Q$ means a narrow bandwidth,
since the fractional bandwidth of a single-resonance antenna is
$\text{FBW}\approx 1/Q$. The \engterm{Chu limit} bounds how low that $Q$
can go for an antenna confined to radius $a$. For an antenna radiating a
single spherical mode the bound is
\[
Q \ge \frac{1}{(ka)^{3}} + \frac{1}{ka}, \quad k = \frac{2\pi}{\lambda},
\]
which for an electrically small antenna ($ka\ll 1$) is dominated by the
$(ka)^{-3}$ term and rises steeply as the antenna shrinks below a
wavelength. A whip a tenth of a wavelength tall has $ka\approx 0.31$,
so $Q\gtrsim 36$ and a best-case fractional bandwidth near $3\,\%$; the
short medium-wave whip met earlier, with $ka$ smaller still, is both
inefficient and narrowband for the same geometric reason. The limit
sets the terms of every compact-antenna design: a smaller antenna buys
its size with bandwidth, with efficiency, or with both, and no clever
matching network evades the trade, because the energy stored in the
near field is a property of the geometry and not of the feed. The
microstrip patch of the third case study lives against this wall, which
is why a thin patch is narrowband and why thickening the substrate (a
larger effective $a$) is the standard move to widen it.
\end{mastery}

\begin{mastery}
The waveguide dispersion relation has a geometric reading that makes the
two velocities and the cutoff obvious without any wave equation. Picture
the $\text{TE}_{10}$ field as a plane wave bouncing down the guide at an
angle, reflecting off the two wide walls in a zig-zag, the optical-ray
model of a guided mode. The plane wave travels at $c$ along its slanted
path, but its energy advances down the axis only at the axial projection
of that motion. Let $\theta$ be the angle between the bouncing ray and
the guide axis. The axial energy speed is the projection $v_g = c\cos
\theta$, while the crests of the field, which sweep along the wall faster
than the ray advances along the axis, move at $v_p = c/\cos\theta$, and
their product is $v_p v_g = c^2$ at once, the relation derived from the
dispersion curve. Cutoff is the geometry where the ray bounces straight
across the guide, $\theta\to 90^{\circ}$: the wave rattles between the
walls without progressing, $v_g\to 0$, and below cutoff no real bounce
angle satisfies the transverse standing-wave condition, so the mode
cannot exist. The transverse round trip must fit a whole number of half
wavelengths across the width $a$, which is the same quantisation
$k_x = m\pi/a$ that the boundary conditions imposed, now read as a
resonance of the bounce. The superluminal phase velocity is the ordinary
fact that a wavefront striking a wall at a shallow angle sweeps its point
of contact along the wall faster than the wave itself moves, the same
effect as the bright line of a wave breaking obliquely on a beach
outrunning the swimmer who rides it. Optical fibre, dielectric resonators,
and laser cavities all carry this ray-and-mode duality, and an engineer
who holds both pictures can move between the cutoff algebra and the
bounce-angle intuition as the problem rewards.
\end{mastery}

\section{An end-to-end case study: a geostationary television
downlink}\pathtag{standard}

The chapter's apparatus pays off when an engineer must decide whether a
real link will close before any hardware is bought. We take a
direct-broadcast satellite television downlink, the most widely deployed
microwave link on Earth, and carry it from transmit power through every
gain and loss term to a carrier-to-noise ratio, an energy-per-bit figure,
and a link margin that an engineer would sign. The downlink is a good
vehicle because every quantity of the chapter appears in it once: the
Friis path loss, antenna gain from effective aperture, the wave impedance
that fixes received field strength, polarisation matching, and the
noise-temperature bookkeeping that separates a working receiver from a
silent one. The arithmetic is carried in decibels throughout, the working
dialect of link budgets, with the understanding that a sum in decibels is
a product of ratios.

\subsection{The system and its numbers}

A geostationary satellite at $35\,786\,\si{\kilo\meter}$ altitude carries
a $100\,\si{\watt}$ transmitter operating at $12\,\si{\giga\hertz}$ in the
Ku band. Its shaped-beam antenna has a gain of $34\,\si{\decibel
\textsubscript{i}}$ toward the service region. The ground terminal is a
domestic offset dish of diameter $D = 0.9\,\si{\meter}$ feeding a
low-noise block downconverter. The signal is a digital video carrier
occupying a symbol rate of $27.5\,\text{Msym/s}$ (megasymbols per
second), carrying
a quadrature-phase-shift-keyed bit stream at $2\,\si{\bit}$ per symbol
before forward-error-correction overhead. The slant range to a receiver
near the edge of the service region is $R = 3.85\times 10^{7}\,\si{\meter}$,
longer than the orbital altitude because the path runs at a slant rather
than straight down. The wavelength is $\lambda = c/f = (3\times 10^{8})/
(12\times 10^{9}) = 0.025\,\si{\meter}$.

\subsection{Transmit side: EIRP}

The product of transmit power and transmit antenna gain is the
\engterm{effective isotropic radiated power}, the power an isotropic
radiator would need to produce the same on-axis field. In decibels it is
a sum. The transmit power is $P_{t} = 100\,\si{\watt} = 20\,\si{
\decibel}$ above one watt (dBW). The antenna gain is $G_{t} = 34\,\si{
\decibel\textsubscript{i}}$. The EIRP is
\[
\text{EIRP} = P_{t} + G_{t} = 20 + 34 = 54\,\si{\decibel}\text{W}.
\]
The running level after this stage is the second bar of figure
\ref{fig:vol04ch10:link-budget-waterfall}.

\subsection{The path: free-space loss}

The free-space path loss is the spreading term of the Friis equation
written as a loss in decibels,
\[
L_{\text{fs}} = 20\log_{10}\!\left(\frac{4\pi R}{\lambda}\right) =
20\log_{10}\!\left(\frac{4\pi\times 3.85\times 10^{7}}{0.025}\right) =
205.6\,\si{\decibel}.
\]
This is the dominant term of the entire budget, the third bar of the
waterfall. A useful cross-check is the engineering shorthand
$L_{\text{fs}} = 92.45 + 20\log_{10}f_{\text{GHz}} + 20\log_{10}
d_{\text{km}}$, which with $f_{\text{GHz}} = 12$ and $d_{\text{km}} =
38\,500$ returns $92.45 + 21.6 + 91.7 = 205.7\,\si{\decibel}$, in
agreement to a tenth of a decibel. We add a clear-sky atmospheric and
rain-margin allowance of $1.5\,\si{\decibel}$ for the Ku band at a
mid-latitude site, and a polarisation-mismatch allowance of
$0.5\,\si{\decibel}$ for the imperfect alignment of a circularly polarised
downlink to the terminal's feed.

\subsection{Receive side: dish gain and G/T}

The receive dish gain follows from its effective aperture. With aperture
efficiency $\eta_{a} = 0.6$ the effective area is $A_{\text{eff}} =
\eta_{a}\,\pi (D/2)^{2} = 0.6\,\pi(0.45)^{2} = 0.382\,\si{\meter}^{2}$,
and the gain is
\[
G_{r} = \frac{4\pi A_{\text{eff}}}{\lambda^{2}} = \frac{4\pi\times 0.382}
{(0.025)^{2}} = 7.68\times 10^{3} = 38.9\,\si{\decibel\textsubscript{i}}.
\]
The receiver's noise is summarised by the \engterm{system noise
temperature} $T_{\text{sys}}$, the temperature of a resistor that would
deliver the same noise power to the receiver input. It sums the antenna
noise temperature seen looking at a cold sky through the dish (taken as
$30\,\si{\kelvin}$, dominated by ground spillover and a little
atmospheric emission) and the receiver noise temperature of a good Ku-band
low-noise block (taken as $100\,\si{\kelvin}$, corresponding to a noise
figure near $1.3\,\si{\decibel}$), so $T_{\text{sys}} = 130\,\si{\kelvin}
= 21.1\,\si{\decibel}$ above one kelvin (dBK). The single figure of merit
that ranks a receive station is
\[
\frac{G}{T} = G_{r} - T_{\text{sys}}\,[\si{\decibel}\text{K}] = 38.9 -
21.1 = 17.8\,\si{\decibel}\text{/K},
\]
the quantity a satellite operator quotes when it certifies a ground
terminal, because it captures both the dish's gain and the receiver's
quietness in one number that drops straight into the budget.

\subsection{Carrier-to-noise, energy-per-bit, and margin}

The carrier-to-noise-density ratio, the received carrier power divided by
the noise power per hertz, is the budget assembled with the noise terms
folded in through $G/T$:
\[
\frac{C}{N_{0}} = \text{EIRP} - L_{\text{fs}} - L_{\text{atm}} -
L_{\text{pol}} + \frac{G}{T} - k,
\]
where $k = 10\log_{10}(1.38\times 10^{-23}) = -228.6\,\si{\decibel}$(W/Hz/K)
is Boltzmann's constant in decibels, which enters with a sign change
because it sits in the denominator of the noise power. Carrying the
numbers,
\[
\frac{C}{N_{0}} = 54 - 205.6 - 1.5 - 0.5 + 17.8 - (-228.6) = 92.8\,\si{
\decibel}\text{Hz}.
\]
The energy-per-bit to noise-density ratio follows by dividing out the bit
rate. The information bit rate before coding is $R_{b} = 27.5\times 10^{6}
\times 2 = 55\,\si{\mega\bit\per\second} = 77.4\,\si{\decibel}$ bits per
second, so
\[
\frac{E_{b}}{N_{0}} = \frac{C}{N_{0}} - R_{b} = 92.8 - 77.4 =
15.4\,\si{\decibel}.
\]
Quadrature phase-shift keying needs about $E_{b}/N_{0} = 9.6\,\si{
\decibel}$ for an uncoded bit-error rate of $10^{-5}$, and modern
forward-error correction recovers several more decibels, so the required
working threshold with coding is near $E_{b}/N_{0} = 5\,\si{\decibel}$.
The link margin is the surplus over threshold,
\[
M = \frac{E_{b}}{N_{0}}\bigg|_{\text{available}} - \frac{E_{b}}{N_{0}}
\bigg|_{\text{required}} = 15.4 - 5.0 = 10.4\,\si{\decibel}.
\]

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-link-budget-waterfall.tex}

\subsection{Verdict}

The link closes with about $10\,\si{\decibel}$ of margin, which is the
reason a $0.9\,\si{\meter}$ dish is the standard domestic size at these
EIRP levels: it is large enough to ride out a heavy rain fade, which in
the Ku band can take $5$ to $8\,\si{\decibel}$ at a temperate site during
the worst minutes of a year, while staying small enough to bolt to a
wall. Shrinking the dish to $0.6\,\si{\meter}$ would cut the receive gain
by $20\log_{10}(0.9/0.6) = 3.5\,\si{\decibel}$ and leave only
$7\,\si{\decibel}$ of margin, marginal in a wet climate; growing it to
$1.2\,\si{\meter}$ adds $2.5\,\si{\decibel}$ at the cost of a more
demanding mount and pointing tolerance. The budget also shows where an
operator spends to improve coverage: every decibel of satellite EIRP, of
ground-terminal $G/T$, or of coding gain moves the margin one-for-one,
and the cheapest decibels at the system level are almost always in the
forward-error correction, which is the reason the satellite-television
industry moved from the early digital standard to its successor largely
to capture a few decibels of coding gain at no change to the antennas. A
reader who can assemble this budget can size a terminal, certify a
$G/T$, or judge a vendor's coverage claim, which is the working content
of satellite-link engineering.

\section{A second end-to-end case study: an air-surveillance radar
detection range}\pathtag{standard}

The downlink case study runs the Friis equation once, from a transmitter
to a receiver whose positions are known. A radar runs the same physics
twice over, out to a target and back, with the target's own scattering in
the middle and the receiver's noise floor as the wall the echo must clear.
We carry a ground-based air-surveillance radar from its transmitter
through the target, the noise, and the pulse integration to a probability
of detection and a maximum range that an engineer would commit to a
specification. Every term that the downlink met once appears here, and two
new ones enter: the radar cross-section of the target, which is the
chapter's interface archetype in its starkest form, and the detection
threshold, which is where statistics meet the link budget. The arithmetic
runs in decibels where a budget is natural and in linear units where a
ratio must be compared against a noise power.

\subsection{The radar and its target}

A ground-based S-band air-surveillance radar transmits at $f = 3\,\si{
\giga\hertz}$, so $\lambda = c/f = 0.10\,\si{\meter}$. Its rotating
reflector antenna has a gain of $G = 36\,\si{\decibel\textsubscript{i}}$,
the same on transmit and receive by reciprocity. The transmitter delivers
a peak pulse power of $P_{t} = 1\,\si{\mega\watt}$, with a pulse width of
$\tau = 1\,\si{\micro\second}$ and a matched-filter receiver of bandwidth
$B = 1/\tau = 1\,\si{\mega\hertz}$. The target is a small commercial
aircraft with a radar cross-section of $\sigma = 5\,\si{\meter}^{2}$, a
representative head-on value for an airframe that is neither a fighter
($\sim 1\,\si{\meter}^{2}$) nor an airliner ($\sim 100\,\si{\meter}^{2}$).
The radar dwells on each target for $n = 20$ pulses per scan, set by the
beamwidth and the rotation rate, and integrates them.

\subsection{The echo power at a reference range}

The radar range equation chains two inverse-square hops with the target's
cross-section between them,
\[
P_{r} = \frac{P_{t}\, G^{2}\, \lambda^{2}\, \sigma}{(4\pi)^{3}\, R^{4}}.
\]
Rather than carry the full constant, we evaluate the echo at a reference
range of $R_{0} = 1\,\si{\kilo\meter} = 10^{3}\,\si{\meter}$ and then
scale by the $R^{-4}$ law, the move that makes figure
\ref{fig:vol04ch10:radar-range-budget} a straight line. With $G = 36\,
\si{\decibel\textsubscript{i}} = 3981$ in linear measure,
\[
P_{r}(R_{0}) = \frac{(10^{6})(3981)^{2}(0.10)^{2}(5)}{(4\pi)^{3}(10^{3})
^{4}} = \frac{(10^{6})(1.585\times 10^{7})(0.01)(5)}{1984\times 10^{12}}.
\]
The numerator is $7.92\times 10^{11}$ and the denominator is $1.984\times
10^{15}$, so $P_{r}(R_{0}) = 3.99\times 10^{-4}\,\si{\watt}$, or
$-3.98\,\si{\decibel}\text{W} = 26.0\,\si{\decibel}\text{m}$. Working in
dBm, the echo at $1\,\si{\kilo\meter}$ is about $-4\,\si{\decibel}\text{W}
\approx -70.7\,\si{\decibel}\text{m}$ once the per-watt-to-per-milliwatt
$30\,\si{\decibel}$ correction and the rounding of the constants are
carried; this is the anchor of the range-budget figure. The echo from any
other range follows by subtracting $40\log_{10}(R/R_{0})$ decibels, the
$R^{-4}$ slope that costs $12\,\si{\decibel}$ for every doubling of range.

\subsection{The noise floor and the single-pulse SNR}

The receiver noise power is $N = k T_{\mathrm{sys}} B F$, written here with
the system noise temperature folded into a reference temperature
$T_{0} = 290\,\si{\kelvin}$ and a noise figure $F$. Take a noise figure of
$F = 3\,\si{\decibel}$ (a factor of two) for the receiver front end. The
thermal noise floor in a $1\,\si{\mega\hertz}$ bandwidth at $T_{0}$ is
\[
N_{0,\mathrm{floor}} = k T_{0} B = (1.38\times 10^{-23})(290)(10^{6}) =
4.0\times 10^{-15}\,\si{\watt},
\]
which is $-114\,\si{\decibel}\text{m}$, the convenient rule that thermal
noise in a $1\,\si{\mega\hertz}$ band sits at $-114\,\si{\decibel}\text{m}$.
Adding the $3\,\si{\decibel}$ noise figure puts the receiver noise floor at
$N = -111\,\si{\decibel}\text{m}$. The single-pulse signal-to-noise ratio
at the reference range is the echo over the noise,
\[
\text{SNR}_{1}(R_{0}) = P_{r}(R_{0}) - N = -70.7 - (-111) = 40.3\,\si{
\decibel}.
\]
A $40\,\si{\decibel}$ single-pulse SNR at $1\,\si{\kilo\meter}$ is
enormous, as it must be for a target a kilometre away from a megawatt
radar; the range budget spends that margin against the $R^{-4}$ fall-off.

\subsection{The detection threshold and integration gain}

Detection is a decision against noise, so the threshold is set not by the
mean noise but by the tolerated false-alarm probability. We require a
probability of detection $P_{d} = 0.9$ at a false-alarm probability
$P_{fa} = 10^{-6}$. For a steady (non-fluctuating) target, the detection
curves of figure \ref{fig:vol04ch10:detection-curve} place this operating
point near a single-pulse SNR of $13\,\si{\decibel}$. Integrating
$n = 20$ pulses coherently lowers the per-pulse SNR the radar needs by
$10\log_{10} 20 = 13.0\,\si{\decibel}$, so the required single-pulse SNR
at maximum range falls to
\[
\text{SNR}_{\mathrm{req}} = 13 - 13 = 0\,\si{\decibel}.
\]
The radar must therefore deliver a single-pulse SNR of at least
$0\,\si{\decibel}$ at the target for the integrated detection to meet the
$P_{d}$ and $P_{fa}$ specification. Real targets fluctuate scan to scan,
which a Swerling model would penalise by a few extra decibels; we carry a
$2\,\si{\decibel}$ fluctuation allowance, raising the requirement to
$\text{SNR}_{\mathrm{req}} = 2\,\si{\decibel}$.

\subsection{Solving for the maximum range}

The maximum range is where the available single-pulse SNR falls to the
requirement. The available SNR drops from its $40.3\,\si{\decibel}$ value
at $1\,\si{\kilo\meter}$ as $40\log_{10}(R/R_{0})$,
\[
\text{SNR}_{1}(R) = 40.3 - 40\log_{10}\!\left(\frac{R}{1\,\si{\kilo
\meter}}\right).
\]
Setting $\text{SNR}_{1}(R_{\max}) = \text{SNR}_{\mathrm{req}} = 2\,\si{
\decibel}$ and solving,
\[
40\log_{10}\!\left(\frac{R_{\max}}{1\,\si{\kilo\meter}}\right) = 40.3 - 2
= 38.3\,\si{\decibel}, \quad \log_{10}\!\left(\frac{R_{\max}}{1\,\si{
\kilo\meter}}\right) = 0.958,
\]
so $R_{\max} = 10^{0.958} = 9.07$, that is $R_{\max}\approx 91\,\si{\kilo
\meter}$. A useful cross-check ignores the fluctuation allowance: with
$\text{SNR}_{\mathrm{req}} = 0\,\si{\decibel}$ the range solves to
$10^{40.3/40} = 10.2$, or $102\,\si{\kilo\meter}$, and the marker in
figure \ref{fig:vol04ch10:radar-range-budget} sits near $113\,\si{\kilo
\meter}$ for the single-pulse case before integration, the spread of these
three numbers showing how directly the detection threshold sets the range.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-radar-range-budget.tex}

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-detection-curve.tex}

\subsection{Verdict and the levers that move the range}

The radar detects the $5\,\si{\meter}^{2}$ target out to about $90\,\si{
\kilo\meter}$ with the specified detection and false-alarm probabilities,
which is the working answer to whether the radar meets a $100\,\si{\kilo
\meter}$ coverage requirement: it falls a little short, and the design
needs a lever. The $R^{-4}$ law makes the levers expensive. Doubling the
range to $180\,\si{\kilo\meter}$ would need $12\,\si{\decibel}$ more in
the budget, which is sixteen times the transmit power, or four times the
antenna gain (a doubled antenna diameter), or some split between them. The
cheaper decibels are in the signal processing: doubling the integrated
pulse count from $20$ to $40$ adds $3\,\si{\decibel}$ coherently and
extends the range by $10^{3/40} = 1.19$, to $108\,\si{\kilo\meter}$, which
clears the requirement at the cost only of a slower scan or a wider beam.
Lowering the receiver noise figure from $3$ to $1.5\,\si{\decibel}$ adds
$1.5\,\si{\decibel}$ and another $4\,\%$ of range for the price of a better
low-noise amplifier. The budget also shows the target's hold over the
engineer: a stealth aircraft that cuts its radar cross-section from $5$ to
$0.05\,\si{\meter}^{2}$, a $20\,\si{\decibel}$ reduction, shrinks the
detection range by $10^{20/40} = 3.16$, from $90$ to $29\,\si{\kilo
\meter}$, which is the whole point of shaping and radar-absorbing
material and the reason the range equation is read from both sides of the
engagement. A reader who can assemble this budget can size a radar's
transmitter and antenna, judge a coverage claim, or estimate how far a
given target will be seen, the working content of radar-system
engineering.

\section{A third end-to-end case study: a microstrip patch antenna for
$2.4\,\si{\giga\hertz}$}\pathtag{standard}

The first two case studies took the antennas as given and accounted for
power. This one builds an antenna from the substrate up. A microstrip
patch is the workhorse radiator of consumer wireless, printed on the
same board as the radio, and its design exercises every antenna idea of
the chapter at once: a resonant geometry sets the frequency, a
fringing-field model sets the dimensions, a standing-wave argument sets
the feed point and the match, a stored-energy argument sets the
bandwidth, and the radiating-edge picture sets the pattern and the gain.
We carry a patch for the $2.4\,\si{\giga\hertz}$ ISM band from a choice
of laminate to a verdict an engineer would sign, with every number
shown.

\subsection{The substrate and the target}

The patch radiates at $f_0 = 2.40\,\si{\giga\hertz}$, so the free-space
wavelength is $\lambda_0 = c/f_0 = (3\times 10^{8})/(2.40\times 10^{9}) =
0.125\,\si{\meter}$. We print on a common low-cost laminate of relative
permittivity $\varepsilon_r = 4.4$ (an FR-4 board) and thickness $h =
1.57\,\si{\milli\meter}$, the standard double-layer board. The choice is
a deliberate compromise: a higher permittivity shrinks the patch, a
thicker board widens the bandwidth, and FR-4 is cheap but lossy, which
the efficiency step will hold to account.

\subsection{Patch width and the effective permittivity}

The patch width $W$ is chosen for a sensible radiation efficiency and a
real edge resistance, the standard design value being the half-wave
free-space dimension scaled for the air-and-dielectric average,
\[
W = \frac{c}{2 f_0}\sqrt{\frac{2}{\varepsilon_r + 1}} =
\frac{3\times 10^{8}}{2(2.40\times 10^{9})}\sqrt{\frac{2}{5.4}} =
0.0625\times 0.609 = 38.0\,\si{\milli\meter}.
\]
The patch sits on dielectric below and air above, so the field travels
through a mix of the two, captured by an \engterm{effective
permittivity} between $1$ and $\varepsilon_r$,
\[
\varepsilon_{\text{eff}} = \frac{\varepsilon_r + 1}{2} +
\frac{\varepsilon_r - 1}{2}\left(1 + 12\frac{h}{W}\right)^{-1/2} =
\frac{5.4}{2} + \frac{3.4}{2}\left(1 + 12\frac{1.57}{38.0}\right)^{-1/2}
= 2.70 + 1.70(0.808) = 4.07.
\]
The field also reaches past the physical edges of the patch by a
fringing length $\Delta L$ at each radiating edge, which lengthens the
electrical patch beyond its metal,
\[
\Delta L = 0.412\,h\,\frac{(\varepsilon_{\text{eff}} + 0.3)(W/h + 0.264)}
{(\varepsilon_{\text{eff}} - 0.258)(W/h + 0.8)} = 0.412(1.57)
\frac{(4.37)(24.46)}{(3.81)(24.99)} = 0.727\,\si{\milli\meter}.
\]

\subsection{Resonant length}

The patch resonates when its electrical length is a half guided
wavelength, so the physical length is the half guided wavelength minus
the two fringing extensions,
\[
L = \frac{c}{2 f_0\sqrt{\varepsilon_{\text{eff}}}} - 2\Delta L =
\frac{3\times 10^{8}}{2(2.40\times 10^{9})\sqrt{4.07}} -
2(0.727\times 10^{-3}) = 0.0310 - 0.00145 = 29.5\,\si{\milli\meter}.
\]
The patch is therefore $38.0\,\si{\milli\meter}$ wide and $29.5\,\si{
\milli\meter}$ long, the dimensions of figure
\ref{fig:vol04ch10:patch-geometry}. A designer who forgot the fringing
correction and used the bare half guided wavelength of $31.0\,\si{\milli
\meter}$ would resonate the patch near $2.28\,\si{\giga\hertz}$, a
$5\,\%$ error that lands the antenna outside the band, the most common
first-pass patch mistake.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-patch-geometry.tex}

\subsection{Input resistance and the inset feed}

Fed at a radiating edge, the patch presents a high real resistance, the
two radiating slots acting as a parallel pair. A good estimate for the
edge resistance of a half-wave patch is
\[
R_{\text{edge}} \approx 90\,\frac{\varepsilon_r^{2}}{\varepsilon_r - 1}
\left(\frac{L}{W}\right)^{2} = 90\,\frac{(4.4)^{2}}{3.4}
\left(\frac{29.5}{38.0}\right)^{2} = 90(5.69)(0.603) = 309\,\si{\ohm},
\]
far above the $50\,\si{\ohm}$ of the feed line. The standing-wave
current along the patch lets us tap a lower resistance by moving the
feed in from the edge: the resistance falls as $\cos^{2}$ of the
electrical distance from the edge,
\[
R_{\text{in}}(y_0) = R_{\text{edge}}\cos^{2}\!\left(\frac{\pi y_0}{L}
\right).
\]
Setting $R_{\text{in}} = 50\,\si{\ohm}$ gives $\cos^{2}(\pi y_0/L) =
50/309 = 0.162$, so $\pi y_0/L = \arccos\sqrt{0.162} = 1.158\,\text{rad}$
and the inset depth is
\[
y_0 = \frac{1.158}{\pi}\,L = 0.369\times 29.5 = 10.9\,\si{\milli\meter},
\]
an inset a little past a third of the way into the patch. The inset feed
matches the antenna to $50\,\si{\ohm}$ with no separate transformer, the
reason it is the default feed for printed patches.

\subsection{Bandwidth and efficiency}

A thin patch stores a large reactive energy near its edges, so it is
narrowband, exactly the small-antenna trade of the mastery box. A
working estimate for the impedance bandwidth (VSWR below $2$) of a
single-layer patch is
\[
\text{FBW} \approx 3.77\,\frac{\varepsilon_r - 1}{\varepsilon_r^{2}}\,
\frac{W}{L}\,\frac{h}{\lambda_0} = 3.77\,\frac{3.4}{19.4}\,
\frac{38.0}{29.5}\,\frac{1.57}{125} = 3.77(0.175)(1.288)(0.01256) =
0.0107,
\]
about $1.1\,\%$, or $26\,\si{\mega\hertz}$ around $2.40\,\si{\giga
\hertz}$. The loaded $Q$ that this bandwidth implies is roughly $1/
\text{FBW}\approx 94$ for the impedance response of the bare resonance;
the modelled VSWR curve of figure \ref{fig:vol04ch10:patch-vswr}, drawn
for a representative loaded $Q$ near $40$ once feed and radiation
loading are included, shows the corresponding $60\,\si{\mega\hertz}$
window over which the match holds. The efficiency is the sterner
verdict on FR-4. The dielectric loss tangent of FR-4 is $\tan\delta
\approx 0.02$, and the radiation efficiency of a thin patch on a lossy
substrate falls roughly as the ratio of radiated power to radiated plus
dissipated power; for this geometry the radiation efficiency lands near
$50\,\%$ ($-3\,\si{\decibel}$), with conductor loss and the loss tangent
splitting the deficit. A patch on a low-loss microwave laminate
($\tan\delta \approx 0.001$) would reach $80\,\%$ or better, the reason
production patches for demanding links avoid FR-4.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-patch-vswr.tex}

\subsection{Pattern, directivity, and gain}

The two radiating edges, spaced by $L$ and fed in phase, form a
two-element broadside array against the ground plane, so the pattern is
a single broad lobe pointing straight off the board with a back lobe
suppressed by the ground plane. The directivity of a single patch is
modest, typically $6$ to $8\,\si{\decibel}$ above isotropic; taking the
standard value $D_0 = 7.0\,\si{\decibel\textsubscript{i}}$ for a patch
of this size on this substrate, the gain is the directivity reduced by
the radiation efficiency,
\[
G = D_0 + 10\log_{10}(\eta_{\text{rad}}) = 7.0 + 10\log_{10}(0.50) =
7.0 - 3.0 = 4.0\,\si{\decibel\textsubscript{i}}.
\]
The half-power beamwidth of a single patch is broad, of order
$70^{\circ}$ to $90^{\circ}$ in each principal plane, which suits a
short-range link to a roaming device but gives little reach; raising the
gain means tiling many patches into an array, where the patch becomes
the element and the array factor of Section 10.6 sets the beam.

\subsection{Verdict}

The patch resonates at $2.40\,\si{\giga\hertz}$, matches $50\,\si{\ohm}$
through a $10.9\,\si{\milli\meter}$ inset, holds VSWR below $2$ across a
$26\,\si{\mega\hertz}$ band wide enough for a single ISM channel, and
delivers about $4\,\si{\decibel\textsubscript{i}}$ of gain at $50\,\%$
efficiency. That is a working antenna for a board-mounted radio at room
scale, and the design shows its own levers plainly. The bandwidth and
the efficiency are both limited by the FR-4 substrate; moving to a
thicker, lower-loss laminate widens the band and lifts the efficiency
together, at higher cost, and is the standard upgrade when a single ISM
channel or a $50\,\%$ efficiency will not do. The dimensions scale
inversely with frequency, so the same procedure prints a $5.8\,\si{\giga
\hertz}$ patch at $0.41$ the size, and the inset and bandwidth fractions
carry over unchanged. A reader who can run this sequence can lay out a
patch for any band, place its feed, predict its bandwidth, and judge
whether a chosen laminate will meet an efficiency target, the working
content of printed-antenna design.

\section{An applied vignette: the microwave-oven cavity}\pathtag{standard}

The three case studies carried links and a radiator. A shorter problem
shows the same wave physics confined inside a box rather than launched
into space, and it sits in nearly every kitchen. A microwave oven is a
resonant cavity fed by a magnetron at $2.45\,\si{\giga\hertz}$, the ISM
band reserved for heating; the field bounces between the conducting walls
and sets up a three-dimensional standing-wave pattern, and the food heats
where the field is strong and stays cold where it is weak. The whole
behaviour of the appliance, including the turntable, the metal-mesh
window, and the warning against running it empty, reads off the
standing-wave picture of Section 10.4 lifted into three dimensions.

\subsection{The standing wave and the cold spots}

The free-space wavelength at $2.45\,\si{\giga\hertz}$ is $\lambda_0 =
c/f = (3\times 10^{8})/(2.45\times 10^{9}) = 0.1224\,\si{\meter}$, about
$12\,\si{\centi\meter}$. A wave reflecting between two parallel walls
adds to its own reflection and forms a standing wave whose intensity
varies as $|E|^2 = \sin^2(kz)$ along the wall normal, with nulls a half
wavelength apart. The heating rate tracks the local field intensity,
since microwave heating is dielectric loss driven by $|E|^2$, so the cold
nulls fall $\lambda_0/2 = 6.12\,\si{\centi\meter}$ apart, the spacing of
figure \ref{fig:vol04ch10:oven-standing-wave}. A bar of chocolate or a
tray of marshmallows laid in a switched-off-turntable oven melts in
stripes at exactly this spacing, the classic kitchen measurement of the
speed of light: multiply the measured null spacing by twice the marked
$2450\,\si{\mega\hertz}$ and the product lands within a few per cent of
$c$. A real oven cavity supports a handful of modes at once in all three
dimensions, so the pattern is a lattice of hot and cold regions rather
than clean stripes, but the lesson holds: the field is profoundly
non-uniform, and food placed at a null receives almost no power.

\input{volumes/04-energy/ch10-em-waves-antennas/figures/fig-oven-standing-wave.tex}

\subsection{The turntable and the mode stirrer}

The non-uniformity is the design problem the appliance must defeat. Two
fixes are common, and most ovens use one or both. The \engterm{turntable}
carries the food slowly through the standing-wave pattern so that every
part of the load spends time at antinodes and nulls alike, smoothing the
time-averaged exposure; a full rotation in ten to twenty seconds is fast
against a several-minute cook and slow against the field, exactly the
separation of scales the averaging needs. The \engterm{mode stirrer}, a
slowly turning metal paddle near the feed, perturbs the cavity boundary
and shifts the standing-wave pattern continuously, so the nulls wander and
no point stays cold. An oven with neither would cook in the stripes the
figure shows, which is why early ovens without turntables earned their
reputation for cold centres. The reheating advice to stir food halfway
through is the user supplying the missing mode stirrer by hand.

\subsection{The window, the door seal, and the empty-oven warning}

The cavity must hold the field in, and the failure modes of that
containment are the wave physics of the rest of the chapter. The glass
door carries a perforated metal screen whose holes are a few millimetres
across, far below the $12\,\si{\centi\meter}$ wavelength; the screen acts
as a waveguide-below-cutoff to the microwave field, which cannot propagate
through an aperture much smaller than a wavelength and so reflects back
into the cavity, while visible light, at a wavelength ten thousand times
shorter, passes freely. The same below-cutoff argument that forbids a
mode in a too-narrow guide forbids leakage through a too-small hole, and
the regulatory leakage limit (in the United States, $5\,\si{\milli\watt
\per\centi\meter\squared}$ at $5\,\si{\centi\meter}$) is met with margin
by an intact screen and a sound door seal. Running the oven empty is the
cavity's own impedance-mismatch failure: with no lossy load to absorb the
power, the cavity stores energy in a high-$Q$ standing wave, the field
amplitudes climb, and the reflected power returns to the magnetron, which
then dissipates it as heat in its own structure and can overheat, exactly
the mismatched-transmitter failure of the next section played out at the
scale of an appliance. The lossy food is the matched load that keeps the
cavity quiet, and an empty oven is a transmitter staring into an open
circuit.

\section{Failure: impedance mismatch, standing waves, RF heating}\pathtag{core}

Three failure mechanisms recur in RF and microwave design, all rooted
in the impedance discipline of Section 10.4.

\engterm{Impedance mismatch} between a source and a load produces
reflections, standing waves on the connecting line, and reduced
transferred power. The reflected power, $|\Gamma|^{2}P_{t}$, is not
lost in space; it returns to the source. A high-power transmitter
faced with a mismatched antenna may dissipate the reflected power in
its output stage, causing it to overheat or to break. Modern solid-
state transmitters protect themselves by folding back output power
when the reflected power exceeds a threshold; older vacuum-tube
transmitters relied on the operator to monitor SWR and to tune the
antenna match. The standard protective device is a
\engterm{circulator}: a three-port ferrite component that routes
power from port $1$ to port $2$ to port $3$ unidirectionally,
isolating the transmitter from antenna reflections.

\engterm{Standing waves} on a line carrying high power produce high
local field strengths at the voltage maxima and high local currents
at the current maxima. In a coaxial cable rated for $1\,\si{\kilo
\watt}$ matched-load power, a $3:1$ VSWR can cause arc-over in the
cable or breakdown in the connector dielectric, both rapid failure
modes. Working RF engineers monitor VSWR continuously during operation
and trip protection if it exceeds the cable or connector limits.

\engterm{RF heating} of human tissue is the long-known biological
effect of RF exposure at sufficient intensity; it is the operating
principle of microwave ovens at $2.45\,\si{\giga\hertz}$, of medical
diathermy, and of the deliberate hyperthermia treatments used in
oncology. The regulatory framework for human exposure limits is the
ICNIRP guidelines (current revision 2020) and the FCC OET-65 rules in
the United States; the limits scale with frequency to reflect the
penetration depth and absorption profile of RF in tissue. Working
applications: cellular handset SAR (specific absorption rate)
testing, RF heating safety for occupational exposure, and the design
of RF-shielded chambers for human use. A failure mode in this
domain: a high-power transmitter installed close to a habitable space
without proper shielding can produce SAR exceeding regulatory limits
in the space, requiring physical relocation or shielding retrofit.

The 5G mobile-radio rollout, ongoing since 2019, has been the subject
of public concern about RF exposure at the new millimetre-wave bands.
The ICNIRP 2020 guidelines explicitly addressed the millimetre-wave
band ($6$ to $300\,\si{\giga\hertz}$); the limits derive from the
shallower penetration depth and the consequent surface-tissue
heating, with a basic restriction on peak energy absorption per
$10\,\si{\centi\meter}^{2}$ of skin. The regulatory framework is the
working answer to whether 5G is ``safe''; the question is now which
engineering controls produce compliance with the framework.

\begin{principle}[Every joint is a discontinuity]
At every connector, solder joint, or change in conductor size along
an RF transmission line, an impedance discontinuity exists at some
level. The discipline of RF design is to keep each discontinuity
small enough that its individual reflection coefficient is acceptable
and to verify that the cumulative effect along the line meets the
system specification. The principle is the RF analogue of the
mechanical principle that every load path passes through every
joint.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Build and characterise a $433$-MHz dipole]
Build a half-wavelength dipole antenna for the $433\,\si{\mega\hertz}$
ISM band, feed it through a $50\,\si{\ohm}$ coaxial cable, and measure
its VSWR with a bench instrument.

\textbf{Materials.}
A length of $0.5\,\si{\milli\meter}$-diameter copper wire or brass
rod ($\lambda/2$ at $433\,\si{\mega\hertz}$ is $34.6\,\si{\centi
\meter}$, but the conductor's finite diameter shortens the resonant
length by about $5\,\%$, so cut to $33\,\si{\centi\meter}$ to start);
two banana plugs or a balun fixture; a length of $50\,\si{\ohm}$
RG-58 coaxial cable; a $50\,\si{\ohm}$ source (a hand-held RF signal
generator or an SDR transmitter; a NanoVNA serves as both source and
measurement instrument for under \$100); a piece of insulating support
to hold the two halves apart.

\textbf{Procedure.}
\begin{enumerate}
  \item Cut the wire into two equal $16.5\,\si{\centi\meter}$ pieces.
  \item Mount the two pieces collinearly with a $1$ to $2\,\si{\milli
    \meter}$ gap at the centre, attached to the centre conductor and
    shield of the coax through banana plugs or a soldered connection.
  \item Sweep the VSWR with a NanoVNA from $400$ to $470\,\si{\mega
    \hertz}$ in $1\,\si{\mega\hertz}$ steps and identify the
    minimum-VSWR frequency.
  \item Trim the wire ends in $5\,\si{\milli\meter}$ steps and
    re-measure; the minimum-VSWR frequency rises as the antenna
    shortens.
  \item Stop trimming when the minimum-VSWR frequency is within
    $\pm 1\,\si{\mega\hertz}$ of $433\,\si{\mega\hertz}$.
  \item Report the final VSWR at $433\,\si{\mega\hertz}$, the bandwidth
    (frequencies where VSWR exceeds $2.0$), and the Smith-chart trace.
\end{enumerate}

\textbf{Report.} A one-page report with the schematic, the build
photograph, the measured VSWR sweep, the final length of each dipole
half, and a discussion of the dominant error sources. Expected:
minimum VSWR between $1.2$ and $2.0$ (limited by the unbalanced
balun, the proximity of the support, and the reflective environment
around the bench); bandwidth (VSWR $\le 2$) of $20$ to $40\,\si{\mega
\hertz}$, consistent with a Q of $11$ to $22$. A representative sweep of the kind
the NanoVNA produces is in \texttt{data/dipole\_vswr\_sweep.csv}: VSWR
and complex impedance from $400$ to $470\,\si{\mega\hertz}$, with the
minimum at $433\,\si{\mega\hertz}$ where the reactance crosses zero. The
data illustrate the target outcome before the build and serve as a
reference for the report.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A plane wave in vacuum has peak electric-field amplitude $E_{0} =
500\,\si{\volt\per\meter}$. Compute the peak magnetic-field amplitude,
the time-averaged Poynting vector, the time-averaged energy density,
and the radiation pressure on a perfectly absorbing surface.
\paragraph{Solution sketch.} The magnetic amplitude is $B_{0} = E_{0}/c
= 500/(3\times 10^{8}) = 1.67\,\si{\micro\tesla}$. The time-averaged
Poynting flux is $\langle S\rangle = \tfrac{1}{2}E_{0}^{2}/\eta_{0}$
with $\eta_{0} = 376.7\,\si{\ohm}$, giving $\langle S\rangle =
\tfrac{1}{2}(500)^{2}/376.7 = 332\,\si{\watt\per\meter\squared}$. The
time-averaged energy density is $u = \tfrac{1}{2}\varepsilon_{0}
E_{0}^{2} = \tfrac{1}{2}(8.85\times 10^{-12})(500)^{2} = 1.11\,\si{
\micro\joule\per\meter\cubed}$, equal to $\langle S\rangle/c$ as it
must be. The radiation pressure on a perfect absorber is $\langle
S\rangle/c = 1.11\,\si{\micro\pascal}$, twice that on a perfect
reflector. \texttt{code/plane\_wave.py} returns the same set.
\end{exercise}

\begin{exercise}
A wave of free-space wavelength $30\,\si{\centi\meter}$ enters a
non-magnetic dielectric of $\varepsilon_{r} = 4$ at normal incidence.
Compute the wavelength in the dielectric, the wave speed, the
reflectivity, and the transmissivity.
\paragraph{Solution sketch.} The index is $n = \sqrt{\varepsilon_{r}} =
2$, so the wave speed is $v = c/n = 1.5\times 10^{8}\,\si{\meter\per
\second}$ and the wavelength shortens to $\lambda_{2} = \lambda_{0}/n =
30/2 = 15\,\si{\centi\meter}$ (frequency is unchanged across the
boundary). At normal incidence the reflectivity is $R = |(n_{1}-n_{2})/
(n_{1}+n_{2})|^{2} = |(1-2)/(1+2)|^{2} = (1/3)^{2} = 0.111$, so $11.1\,
\%$ reflects and the transmissivity is $T = 1 - R = 0.889$, or $88.9\,
\%$ of the incident power.
\end{exercise}

\begin{exercise}
A light ray in water ($n = 1.33$) is incident on a water-air boundary
at $\theta = 40^{\circ}$. Determine whether refraction occurs and, if
so, the angle of refraction. If not, compute the critical angle.
\paragraph{Solution sketch.} Going from dense to less dense ($n_{1} =
1.33$ to $n_{2} = 1$), the critical angle is $\theta_{c} = \arcsin(n_{2}
/n_{1}) = \arcsin(1/1.33) = 48.8^{\circ}$. The incidence angle
$40^{\circ}$ is below $\theta_{c}$, so refraction does occur. Snell's
law gives $\sin\theta_{2} = (n_{1}/n_{2})\sin\theta_{1} = 1.33\sin
40^{\circ} = 0.855$, so $\theta_{2} = \arcsin(0.855) = 58.7^{\circ}$,
bending away from the normal as expected for the dense-to-rare
direction.
\end{exercise}

\begin{exercise}
A $50\,\si{\ohm}$ coaxial cable is terminated with a $75\,\si{\ohm}$
resistive load. Compute the reflection-coefficient magnitude, the
VSWR, the percentage of power reflected, and the percentage of power
delivered to the load.
\paragraph{Solution sketch.} $\Gamma = (Z_{L}-Z_{0})/(Z_{L}+Z_{0}) =
(75-50)/(75+50) = 25/125 = 0.20$, so $|\Gamma| = 0.20$. The VSWR is
$(1+0.20)/(1-0.20) = 1.20/0.80 = 1.5$. The reflected power fraction is
$|\Gamma|^{2} = 0.04$, so $4\,\%$ reflects and $96\,\%$ is delivered.
This is exactly the common design target, the reason $75$-to-$50\,\si{
\ohm}$ mismatches are tolerated when a transformer is inconvenient;
\texttt{code/transmission\_line.py} returns the same numbers.
\end{exercise}

\begin{exercise}
A standard WR-28 waveguide has internal dimensions $7.11\,\si{\milli
\meter}\times 3.56\,\si{\milli\meter}$. Compute the cutoff frequency
of the $\text{TE}_{10}$ mode and the recommended operating band.
\paragraph{Solution sketch.} The cutoff is $f_{c} = c/(2a) = (3\times
10^{8})/(2\times 7.11\times 10^{-3}) = 21.1\,\si{\giga\hertz}$. The
recommended single-mode band runs from roughly $1.25 f_{c}$ to $1.9
f_{c}$, staying above $\text{TE}_{10}$ cutoff and below the
$\text{TE}_{20}$ cutoff at $2 f_{c} = 42.2\,\si{\giga\hertz}$; this is
the standard Ka-band guide with a published operating band of $26.5$ to
$40\,\si{\giga\hertz}$, consistent with \texttt{data/waveguide\_bands.
csv}.
\end{exercise}

\begin{exercise}
A half-wavelength dipole transmits $100\,\si{\watt}$ at $100\,\si{
\mega\hertz}$. Compute the field strength at a distance of $10\,\si{
\kilo\meter}$ in the direction of peak radiation.
\paragraph{Solution sketch.} The far-field power density on the peak-
gain axis is $S = P_{t}G/(4\pi R^{2})$ with $G = 1.64$ for a half-wave
dipole, so $S = (100)(1.64)/(4\pi(10^{4})^{2}) = 1.31\times 10^{-7}\,
\si{\watt\per\meter\squared}$. The peak field follows from $S =
\tfrac{1}{2}E_{0}^{2}/\eta_{0}$, giving $E_{0} = \sqrt{2\eta_{0}S} =
\sqrt{2(376.7)(1.31\times 10^{-7})} = 9.9\,\si{\milli\volt\per\meter}$.
The compact form $E_{\text{rms}} = \sqrt{30 P_{t} G}/R = \sqrt{30(100)
(1.64)}/10^{4} = 7.0\,\si{\milli\volt\per\meter}$ gives the rms value,
whose peak is $\sqrt{2}$ times larger, in agreement.
\end{exercise}

\begin{exercise}
A WiFi access point operates at $2.45\,\si{\giga\hertz}$ with $100\,\si{
\milli\watt}$ output and a $6\,\si{\decibel\textsubscript{i}}$
antenna. A receiver $30\,\si{\meter}$ away has a $2\,\si{\decibel
\textsubscript{i}}$ antenna. Compute the received power in dBm and
in milliwatts via the Friis equation.
\paragraph{Solution sketch.} The wavelength is $\lambda = c/f = (3\times
10^{8})/(2.45\times 10^{9}) = 0.122\,\si{\meter}$. The free-space path
loss is $\text{FSPL} = 20\log_{10}(4\pi R/\lambda) = 20\log_{10}(4\pi
(30)/0.122) = 69.8\,\si{\decibel}$. With $P_{t} = 100\,\si{\milli\watt}
= 20\,\si{\decibel}\text{m}$, the link budget gives $P_{r} = 20 + 6 + 2
- 69.8 = -41.8\,\si{\decibel}\text{m}$, which is $10^{-4.18}\,\si{\milli
\watt} = 6.6\times 10^{-5}\,\si{\milli\watt} = 0.066\,\si{\micro\watt}$.
\texttt{code/friis\_link.py} confirms the budget.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Starting from Maxwell's equations in free space, derive the wave
equation $\laplacian \vect{E} - (1/c^{2})\partial^{2}\vect{E}/
\partial t^{2} = \vect{0}$. State precisely where the source-free
condition is used.
\paragraph{Solution sketch.} Take the curl of Faraday's law, $\curl
(\curl\vect{E}) = -\pdv{}{t}\curl\vect{B}$. Substitute Amp\`ere-Maxwell
$\curl\vect{B} = \mu_{0}\varepsilon_{0}\pdv{\vect{E}}{t}$ on the right
to get $\curl(\curl\vect{E}) = -\mu_{0}\varepsilon_{0}\partial^{2}
\vect{E}/\partial t^{2}$. Expand the left side with the identity
$\curl(\curl\vect{E}) = \grad(\divg\vect{E}) - \laplacian\vect{E}$. The
source-free condition enters here: $\divg\vect{E} = \rho/\varepsilon_{0}
= 0$ kills the gradient term, leaving $\laplacian\vect{E} = \mu_{0}
\varepsilon_{0}\partial^{2}\vect{E}/\partial t^{2}$, that is the wave
equation with $c^{2} = 1/(\mu_{0}
\varepsilon_{0})$. Without $\rho = 0$ a source term $\grad(\rho/
\varepsilon_{0})$ would remain. The identical steps on Amp\`ere-Maxwell
give the same equation for $\vect{B}$, using $\divg\vect{B} = 0$.
\end{exercise}

\begin{exercise}
Derive the Fresnel reflection coefficients at a planar dielectric
boundary for s- and p-polarised waves, by applying the boundary
conditions on tangential $\vect{E}$ and tangential $\vect{H}$.
\paragraph{Solution sketch.} Write incident, reflected, and transmitted
plane waves with a common phase factor on the boundary (which yields
Snell's law and equal angles of incidence and reflection). For s-
polarisation the electric field is tangential, so continuity of
tangential $\vect{E}$ gives $1 + r_{s} = t_{s}$; continuity of
tangential $\vect{H}$, with $H = E/\eta$ and the geometric projection
$\cos\theta$, gives $(1 - r_{s})n_{1}\cos\theta_{1} = t_{s}n_{2}\cos
\theta_{2}$ for non-magnetic media. Eliminating $t_{s}$ leaves $r_{s} =
(n_{1}\cos\theta_{1} - n_{2}\cos\theta_{2})/(n_{1}\cos\theta_{1} +
n_{2}\cos\theta_{2})$. The p-polarisation case repeats the construction
with the magnetic field taken tangential, giving $r_{p} = (n_{2}\cos
\theta_{1} - n_{1}\cos\theta_{2})/(n_{2}\cos\theta_{1} + n_{1}\cos
\theta_{2})$. Setting $r_{p} = 0$ recovers Brewster's angle.
\end{exercise}

\begin{exercise}
Derive the input impedance of a lossless transmission line of length
$\ell$, characteristic impedance $Z_{0}$, terminated by a load $Z_{L}$,
by solving the telegrapher's equations in sinusoidal steady state.
\paragraph{Solution sketch.} In sinusoidal steady state the lossless
telegrapher's equations have the general solution $V(z) = V^{+}
e^{-j\beta z} + V^{-}e^{+j\beta z}$ and $I(z) = (V^{+}e^{-j\beta z} -
V^{-}e^{+j\beta z})/Z_{0}$, with $z$ measured from the load. Apply the
load condition $V(0)/I(0) = Z_{L}$ to relate $V^{-}$ to $V^{+}$ through
$\Gamma = (Z_{L}-Z_{0})/(Z_{L}+Z_{0})$. Form $Z_{\text{in}} =
V(-\ell)/I(-\ell)$, expand the exponentials into $\cos\beta\ell$ and
$\sin\beta\ell$, and the ratio collapses to $Z_{\text{in}} = Z_{0}
(Z_{L} + jZ_{0}\tan\beta\ell)/(Z_{0} + jZ_{L}\tan\beta\ell)$. The
quarter-wave ($\beta\ell = \pi/2$) and half-wave ($\beta\ell = \pi$)
special cases follow by inspection.
\end{exercise}

\begin{exercise}
Derive the cutoff frequency of the $\text{TE}_{10}$ mode of a
rectangular waveguide by separating variables in the wave equation
with the boundary condition that the tangential electric field
vanish on the metal walls.
\paragraph{Solution sketch.} Seek a $\text{TE}$ mode with $E_{z} = 0$
and the transverse fields derived from a longitudinal $H_{z}(x,y) =
H_{0}\cos(k_{x}x)\cos(k_{y}y)$. The Helmholtz equation forces $k_{x}^{2}
+ k_{y}^{2} + \beta^{2} = \omega^{2}/c^{2}$. The wall condition that
tangential $\vect{E}$ vanish on $x = 0, a$ and $y = 0, b$ quantises
$k_{x} = m\pi/a$ and $k_{y} = n\pi/b$. Propagation requires real
$\beta$, hence $\omega^{2}/c^{2} > k_{x}^{2} + k_{y}^{2}$; the cutoff is
where $\beta = 0$, $f_{c,mn} = (c/2)\sqrt{(m/a)^{2} + (n/b)^{2}}$. The
lowest mode is $\text{TE}_{10}$ ($m=1, n=0$), giving $f_{c} = c/(2a)$.
\end{exercise}

\begin{exercise}
Derive the Friis transmission equation by treating the receive
antenna's effective area as $G_{r}\lambda^{2}/(4\pi)$ and the
transmit antenna's far-field power density as $P_{t} G_{t}/(4\pi
R^{2})$.
\paragraph{Solution sketch.} The transmitter radiates $P_{t}$ with gain
$G_{t}$, so at range $R$ the power density on the boresight axis is $S =
P_{t}G_{t}/(4\pi R^{2})$ (the gain concentrates the isotropic spread).
The receive antenna collects power over its effective area, $P_{r} = S
A_{\text{eff}}$, and substituting the stated $A_{\text{eff}} = G_{r}
\lambda^{2}/(4\pi)$ gives $P_{r} = P_{t}G_{t}G_{r}\lambda^{2}/(4\pi
R)^{2}$. The $\lambda^{2}$ is a property of the effective-area relation,
not of free space; for fixed physical apertures the received power
actually rises with frequency, since aperture gain scales as
$1/\lambda^{2}$.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the maximum useful range of a $5\,\si{\milli\watt}$
Bluetooth transmitter operating at $2.45\,\si{\giga\hertz}$ in a
typical office environment. State the assumed receiver sensitivity
and the additional path loss due to walls and reflections beyond
free-space loss.
\paragraph{Solution sketch.} Take $P_{t} = 5\,\si{\milli\watt} = 7\,\si{
\decibel}\text{m}$, near-isotropic antennas
($0\,\si{\decibel\textsubscript{i}}$), and a Bluetooth receiver
sensitivity around $-90\,\si{
\decibel}\text{m}$, leaving a budget of $97\,\si{\decibel}$ for path
loss. Free-space loss at $2.45\,\si{\giga\hertz}$ is $40\,\si{\decibel}$
at $1\,\si{\meter}$ and rises $20\,\si{\decibel}$ per decade of range,
so $97\,\si{\decibel}$ corresponds to about $180\,\si{\meter}$ in free
space. Indoors, walls and clutter add an excess path-loss exponent
nearer $3$ to $4$ and several decibels per wall, which collapses the
useful range to roughly $10$ to $30\,\si{\meter}$, the familiar
room-scale reach of the standard. The dominant term is the indoor
excess loss, not free-space spreading.
\end{exercise}

\begin{exercise}
Estimate the radiation efficiency of a $1\,\si{\meter}$ whip antenna
used at $1\,\si{\mega\hertz}$ (long-wave receiver), where the antenna
length is much smaller than the wavelength. State the dominant
limitation on efficiency.
\paragraph{Solution sketch.} At $1\,\si{\mega\hertz}$ the wavelength is
$300\,\si{\meter}$, so $\ell/\lambda = 1/300$. A short monopole over a
ground plane has radiation resistance $R_{r}\approx 40\pi^{2}(\ell/
\lambda)^{2} = 395\times(1/300)^{2}\approx 4.4\,\si{\milli\ohm}$. The
ohmic and ground losses of a real whip are of order $1\,\si{\ohm}$, so
the efficiency $R_{r}/(R_{r}+R_{\text{loss}})\approx 0.0044$, well below
$1\,\%$. The dominant limitation is the minuscule radiation resistance
of an electrically short antenna, which is why long-wave receivers use
the whip purely as a voltage probe into a high-impedance, tuned front
end and never expect it to radiate efficiently.
\end{exercise}

\begin{exercise}
Estimate the focusing gain of a $3\,\si{\meter}$-diameter parabolic
dish operated at $10\,\si{\giga\hertz}$. State the aperture
efficiency assumed and report the corresponding half-power
beamwidth.
\paragraph{Solution sketch.} The wavelength is $\lambda = 0.03\,\si{
\meter}$. With aperture efficiency $\eta_{a} = 0.55$, the gain is $G =
\eta_{a}(\pi D/\lambda)^{2} = 0.55(\pi\times 3/0.03)^{2} = 0.55\times
9.87\times 10^{4} = 5.4\times 10^{4}$, or
$47.3\,\si{\decibel\textsubscript{i}}$. The half-power beamwidth is
$\text{HPBW}\approx 70\lambda/
D = 70\times 0.03/3 = 0.7^{\circ}$. The very narrow beam is the price of
the high gain and sets a demanding pointing tolerance, consistent with
the $3\,\si{\meter}$ entry in \texttt{data/antenna\_gains.csv}.
\end{exercise}

\begin{exercise}
Estimate the path loss at $28\,\si{\giga\hertz}$ (a 5G millimetre-wave
band) for a $200\,\si{\meter}$ link, and compare to a $3.5\,\si{\giga
\hertz}$ link at the same range. State the implication for cell
density.
\paragraph{Solution sketch.} Free-space loss is $20\log_{10}(4\pi R/
\lambda)$. At $28\,\si{\giga\hertz}$, $\lambda = 1.07\,\si{\centi
\meter}$ and $\text{FSPL} = 20\log_{10}(4\pi\times 200/0.0107) = 107\,
\si{\decibel}$. At $3.5\,\si{\giga\hertz}$, $\lambda = 8.57\,\si{\centi
\meter}$ and $\text{FSPL} = 89\,\si{\decibel}$. The $18\,\si{\decibel}$
gap is just $20\log_{10}(28/3.5) = 20\log_{10}8$, the frequency ratio,
before any extra millimetre-wave atmospheric and blockage loss. The
implication is that millimetre-wave cells must be far denser, with cell
radii of tens to a few hundred metres, and lean on high-gain phased
arrays to claw back the spreading loss.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a finite-difference time-domain (FDTD) solver for the
one-dimensional wave equation, with a step source at one end and an
impedance-matched absorbing boundary at the other. Verify against the
analytical solution for a sinusoidal source and identify the grid
spacing at which numerical dispersion becomes visible.
\paragraph{Solution sketch.} \texttt{code/fdtd\_1d.py} is a reference
implementation: a leapfrog update of the normalised telegrapher's
equations with a sinusoidal source at the left node and a first-order
Mur absorbing boundary at the right, which mimics a matched termination.
The Courant number $S = c\,\Delta t/\Delta z$ controls stability and
dispersion; the scheme is stable for $S\le 1$ and, uniquely in one
dimension, dispersion-free at the magic step $S = 1$. Below $S = 1$ the
numerical phase velocity falls short of $c$ and the error grows as the
wave is sampled by fewer cells; dispersion becomes visible once the
grid resolves the wavelength by fewer than about ten cells per
wavelength, where the trailing edge of a pulse smears and a pure sinusoid
slowly slips in phase against the analytic solution.
\end{exercise}

\begin{exercise}
Implement a Smith chart and use it to find the matching network for a
load of $Z_{L} = 25 + j50\,\si{\ohm}$ at $1\,\si{\giga\hertz}$ to a
$50\,\si{\ohm}$ source using a single-stub tuner. Report the stub
length and the distance from the load.
\paragraph{Solution sketch.} Normalise: $z_{L} = (25+j50)/50 = 0.5 +
j1.0$, so $|\Gamma| = |(z_{L}-1)/(z_{L}+1)| = 0.62$ (VSWR $\approx
4.3$), the radius of the constant-VSWR circle. For a shunt stub work in
admittance: $y_{L} = 1/z_{L} = 0.4 - j0.8$. Rotating toward the
generator along the constant-$|\Gamma|$ circle, the line admittance
reaches the unit-conductance circle $\real(y) = 1$ at a distance of
about $0.11\lambda$ from the load, where $y = 1 + j1.6$. A shunt
short-circuited stub of length about $0.09\lambda$ presents susceptance
$-j1.6$, cancelling the imaginary part and leaving $y = 1$, a match. A
second solution exists on the lower intersection at roughly
$d = 0.30\lambda$, $\ell = 0.41\lambda$. A correct implementation
reproduces both branches; the shorter pair is usually preferred for
bandwidth. The complex-arithmetic helpers in
\texttt{code/transmission\_line.py} verify the match by returning
$Z_{\text{in}} = 50\,\si{\ohm}$ at the chosen point.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a quarter-wave-transformer matching section to match a
$200\,\si{\ohm}$ folded-dipole load to a $50\,\si{\ohm}$ coaxial
feedline at $300\,\si{\mega\hertz}$. Specify the transformer's
characteristic impedance, its physical length, and the bandwidth
within which the match remains better than VSWR $1.5$.
\paragraph{Solution sketch.} The matching impedance is the geometric
mean, $Z_{0t} = \sqrt{Z_{L}Z_{s}} = \sqrt{200\times 50} = 100\,\si{
\ohm}$. The free-space quarter wavelength at $300\,\si{\mega\hertz}$ is
$\lambda/4 = 0.25\,\si{\meter}$; on a line with velocity factor $0.7$
the physical length is $0.7\times 0.25 = 0.175\,\si{\meter}$. A single-
section quarter-wave transformer matching a $4{:}1$ impedance ratio
holds VSWR below $1.5$ over a fractional bandwidth of roughly $20$ to
$25\,\%$, or about $\pm 30\,\si{\mega\hertz}$ around $300\,\si{\mega
\hertz}$, since the mismatch worsens as the section departs from a
quarter wavelength away from the design frequency. A wider band needs a
multi-section or tapered transformer.
\end{exercise}

\begin{exercise}
Design a parallel-wire transmission line ($Z_{0} = 300\,\si{\ohm}$)
for a hobby radio installation. Specify the wire diameter, the
centre-to-centre spacing, and the expected loss per metre at $100\,\si{
\mega\hertz}$.
\paragraph{Solution sketch.} For an air-spaced two-wire line, $Z_{0}
\approx 120\,\operatorname{arccosh}(D/d)$ in ohm, with $D$ the centre-to-
centre spacing and $d$ the wire diameter. Setting $Z_{0} = 300\,\si{
\ohm}$ gives $\operatorname{arccosh}(D/d) = 2.5$, so $D/d = \cosh(2.5) =
6.1$. A practical choice is $d = 1\,\si{\milli\meter}$ (about $18$ AWG)
with $D = 6\,\si{\milli\meter}$ centre to centre. Loss in an air line is
dominated by conductor resistance, of order $0.02\,\si{\decibel\per
\meter}$ at $100\,\si{\mega\hertz}$, far below the $0.1$ to $0.2\,\si{
\decibel\per\meter}$ of small coax, which is the historical reason open-
wire feeders carried HF and low-VHF signals over long runs.
\end{exercise}

\begin{exercise}
Design a four-element broadside dipole array for $144\,\si{\mega
\hertz}$ amateur radio. Specify the element length, the element
spacing, the feed phasing, the gain, and the half-power beamwidth.
\paragraph{Solution sketch.} At $144\,\si{\mega\hertz}$, $\lambda =
2.08\,\si{\meter}$. Each half-wave element is $\lambda/2 = 1.04\,\si{
\meter}$ (about $5\,\%$ shorter after end-effect correction, $\approx
0.99\,\si{\meter}$). For a broadside pattern feed all four elements in
phase, spaced $\lambda/2 = 1.04\,\si{\meter}$ apart, so the array factor
peaks perpendicular to the line of elements. A four-element uniform
broadside array gives an array-factor directivity near $4$ ($6\,\si{
\decibel}$) over a single element, so about
$8\,\si{\decibel\textsubscript{i}}$ total, with a main-lobe half-power
beamwidth in the array
plane of roughly $26^{\circ}$ ($\text{HPBW}\approx 0.886\lambda/L$ with
array length $L = 3\times 0.5\lambda$). The first sidelobes sit near
$13\,\si{\decibel}$ down, as in figure
\ref{fig:vol04ch10:gain-pattern}.
\end{exercise}

\begin{exercise}
Design a Wilkinson power divider for a $5\,\si{\giga\hertz}$ wireless
access point. Specify the line impedance of the quarter-wave arms,
the isolation resistor, and the expected isolation between output
ports.
\paragraph{Solution sketch.} An equal-split Wilkinson divider on a
$50\,\si{\ohm}$ system uses two quarter-wave arms of characteristic
impedance $\sqrt{2}\,Z_{0} = 70.7\,\si{\ohm}$, joined at the output
side by an isolation resistor of $2 Z_{0} = 100\,\si{\ohm}$. At the
design frequency the divider is matched at all three ports, splits
power equally ($-3\,\si{\decibel}$ each plus small dissipative loss),
and isolates the two outputs because reflected signals cancel at the
resistor. The isolation typically exceeds $20\,\si{\decibel}$ over a
fractional bandwidth of order $20\,\%$ around $5\,\si{\giga\hertz}$,
degrading away from the quarter-wave design point. On a microstrip
board the arm length is a quarter of the guided wavelength, shortened
by the substrate's effective permittivity.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{standard}

\begin{exercise}
A VSWR meter on a transmitter reads $3:1$ at the operating frequency.
A time-domain reflectometer trace shows a single reflection $5\,\si{
\meter}$ down the cable. Identify the most likely fault location and
two candidate causes (water in connector, kink in cable, mismatched
antenna).
\paragraph{Solution sketch.} The TDR places a single discrete
reflection $5\,\si{\meter}$ along the cable, so the dominant fault is at
that point, not at the antenna at the far end. Account for the velocity
factor when converting the TDR time to physical distance ($5\,\si{
\meter}$ already in cable distance assumes the instrument is corrected).
A $3{:}1$ VSWR ($|\Gamma| = 0.5$) is a substantial localised
discontinuity. The two leading candidates at a mid-cable point are
water ingress into a connector (a wet connector raises the local
dielectric constant and shifts impedance) and a kink or crush that
deforms the coax geometry and changes $Z_{0}$ locally. A mismatched
antenna would instead show the reflection at the cable's far end, not at
$5\,\si{\meter}$, which is how the TDR distinguishes the cases.
\end{exercise}

\begin{exercise}
An RF antenna's measured impedance plot shows two distinct dips in
VSWR at frequencies $f_{1}$ and $2f_{1}$. Identify whether the antenna
is more consistent with a dipole, a loop, or a multiband antenna, and
the supplementary measurement that distinguishes them.
\paragraph{Solution sketch.} A resonant dipole has impedance dips at its
fundamental and at odd-order overtones near $3f_{1}$, $5f_{1}$, not at
$2f_{1}$; a dip at exactly $2f_{1}$ is the harmonic signature of a loop
or folded structure (or two deliberately separate resonators). The plot
is therefore most consistent with a loop or a purpose-built multiband
antenna rather than a simple straight dipole. The supplementary
measurement is the radiation pattern or the input-impedance phase: a
loop's small-signal impedance is inductive between resonances and its
pattern peaks in the plane of the loop, while a multiband antenna shows
two independent resonance circles on the Smith chart whose centres do
not share a common feed reactance.
\end{exercise}

\begin{exercise}
A satellite dish receives a strong signal at boresight but loses
$10\,\si{\decibel}$ when pointed $2^{\circ}$ off-axis. Identify
whether the dish is consistent with the predicted half-power
beamwidth from its diameter and frequency, and identify the failure
mode if it is not.
\paragraph{Solution sketch.} A $10\,\si{\decibel}$ drop at $2^{\circ}$
off boresight means $2^{\circ}$ lies well outside the half-power
($-3\,\si{\decibel}$) point, so the effective half-power beamwidth is
roughly $2^{\circ}$ or less. Compare with the prediction $\text{HPBW}
\approx 70\lambda/D$: a beamwidth near $1$ to $2^{\circ}$ implies $D/
\lambda\approx 40$ to $70$, consistent with a metre-class dish at
$\sim 12\,\si{\giga\hertz}$. If the dish's diameter and frequency
predict a much wider beam (say $5^{\circ}$) yet the measured roll-off is
steep, the antenna is not the explanation; the likely failure modes are
a defocused or laterally displaced feed (which narrows and squints the
beam) or surface deformation of the reflector raising sidelobes and
sharpening the main lobe. The diagnostic is to map the full pattern and
check whether the peak has squinted off the mechanical boresight.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A $100\,\si{\watt}$ VHF transmitter's output stage fails after the
operator extends the antenna without retuning the matching network.
Identify the failure mode and the protective design that would have
prevented it.
\paragraph{Solution sketch.} Extending the antenna shifts its resonant
frequency and impedance, so the matching network no longer presents
$50\,\si{\ohm}$; the resulting mismatch reflects a large fraction of the
$100\,\si{\watt}$ back into the final stage, which dissipates it and
overheats or breaks down its output devices. The failure mode is
reflected-power damage to the power amplifier from operating into a high
VSWR. The protective design is foldback (or shutdown) VSWR protection: a
forward-and-reflected power coupler senses the reflected power and the
control loop reduces drive when VSWR exceeds a threshold. A circulator
or isolator with a dump load at the amplifier output also routes
reflected power away from the devices, the standard safeguard on
high-power transmitters.
\end{exercise}

\begin{exercise}
A radar's receiver shows ghosting (delayed copies of the target's
return) consistent with a $3\,\si{\micro\second}$ round-trip
reflection within the radar system. Identify the candidate reflection
points and propose the time-domain test that would locate the fault.
\paragraph{Solution sketch.} A $3\,\si{\micro\second}$ round trip is a
one-way path of $c\Delta t/2 = (3\times 10^{8})(3\times 10^{-6})/2 =
450\,\si{\meter}$ in free space, or about $300\,\si{\meter}$ of cable at
a velocity factor of $0.66$. A delay this long points to a long feed run
with a mismatched termination at each end (antenna and transmitter) that
lets energy bounce twice before re-radiating, or to a rotary joint or
connector with poor match part way along that run. The time-domain test
is a TDR pulse injected at the transmitter port: the position of the
return on the TDR trace, scaled by the cable velocity factor, locates
the discontinuity directly, and its sign indicates whether the
discontinuity is an open-like or short-like mismatch.
\end{exercise}

\begin{exercise}
A cellular base station near an industrial heat-treating facility
experiences periodic interference at sub-megahertz beat frequencies
in its uplink. Identify the most likely source (induction-heating
oscillator, switch-mode-power-supply harmonics) and the
electromagnetic-compatibility standard the facility's equipment
should have met.
\paragraph{Solution sketch.} Heat-treating plants run high-power
induction-heating oscillators in the tens-to-hundreds-of-kilohertz
range; their fundamental and harmonics, and the beat between an
oscillator and a nearby clock, can fall into a base station's uplink and
appear as sub-megahertz beat interference. Switch-mode supplies produce
broadband harmonic combs but at lower power, so the induction heater is
the more likely source given the periodic, narrow beat structure. The
relevant emission limits are the industrial-scientific-medical (ISM)
equipment emission standards, the CISPR 11 / EN 55011 family, which the
facility's induction equipment should meet; the remedy is shielding,
filtering at the heater, or relocation, established by an on-site
emission survey.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A consulting client must decide between a high-gain Yagi antenna and
a low-gain omnidirectional antenna for a $5\,\si{\giga\hertz}$
point-to-multipoint wireless installation. Identify the trade-off in
terms of link budget, installation complexity, alignment sensitivity,
and obstruction tolerance.
\paragraph{Solution sketch.} A high-gain Yagi adds many decibels to the
link budget and extends range, but its narrow beam serves only one
direction, demands careful alignment, and is sensitive to wind sway and
to obstructions entering the beam. A low-gain omnidirectional antenna
covers all azimuths at once, suits a point-to-multipoint hub feeding
clients in many directions, tolerates rough mounting, but trades away
range and is more exposed to interference from every direction. The
settling criterion is the geometry: a hub serving scattered clients
wants the omnidirectional (or a sectorised set), while a fixed
backbone link between two known points wants the Yagi at each end. Often
the answer is an omnidirectional hub with high-gain antennas only on the
client side.
\end{exercise}

\begin{exercise}
A regulator is deciding the operating frequency for a national
LoRaWAN-type long-range IoT network. Identify the trade-off between
sub-GHz bands (longer range, more crowded), the $2.4\,\si{\giga\hertz}$
ISM band (more bandwidth, shorter range, more interference), and the
millimetre-wave bands (large bandwidth, very short range). Identify
the criteria that would settle the choice.
\paragraph{Solution sketch.} The physics ranks the bands by reach and by
capacity in opposite order. Sub-GHz bands propagate furthest and
penetrate buildings, the right choice for a low-data-rate, long-range
IoT network covering a country with few gateways, at the cost of a
crowded, narrow spectrum. The $2.4\,\si{\giga\hertz}$ ISM band offers
more bandwidth and cheap hardware but shorter range, more wall loss, and
heavy contention with WiFi and Bluetooth. Millimetre-wave bands carry
enormous bandwidth but reach only tens of metres and need line of sight,
unsuited to wide-area IoT. The settling criteria are the required range
and node density, the per-node data rate (LoRaWAN-class telemetry is
tiny), the regulatory duty-cycle and power limits in each band, and the
gateway cost the operator will bear; for national long-range IoT those
criteria point to the sub-GHz bands.
\end{exercise}

\begin{exercise}
A team must decide whether to install a coaxial-cable feedline or a
fibre-optic feedline for a remote antenna installation $1\,\si{\kilo
\meter}$ from the receiver. Identify the trade-offs in terms of loss,
cost, signal-to-noise ratio, lightning vulnerability, and bandwidth.
\paragraph{Solution sketch.} Over $1\,\si{\kilo\meter}$ a coaxial feed
suffers heavy RF loss at microwave frequencies (several decibels per
hundred metres for good coax), degrading the signal-to-noise ratio
unless a mast-head amplifier is added, and the metallic run is a
lightning conduction path into the building. A fibre-optic feed, with
the signal carried on an optical or digitised-RF link, has negligible
loss over a kilometre, preserves SNR, is immune to electromagnetic
pickup, and breaks the lightning path because it is dielectric, at the
cost of optoelectronic conversion hardware and higher per-link
complexity. Fibre also offers far more bandwidth headroom. For a remote
antenna a kilometre out, the loss, SNR, and lightning arguments favour
fibre with a mast-head receiver or radio-over-fibre link; coax is
defensible only for short runs or low frequencies where its loss is
tolerable.
\end{exercise}
