\chapter{Statics of rigid bodies}
\label{vol03:ch02}

\epigraph{Two magnitudes, whether commensurable or incommensurable,
balance at distances reciprocally proportional to their weights.}{
Archimedes, \emph{On the Equilibrium of Planes}, Book I, Propositions
6 and 7, in the Heath edition \cite[pp.~189--192]{hist:archimedes-equilibrium}.}

\chapmeta{Half-life: durable. Archetypes: balance, failure (joint
mechanics). Exercise target: 40. Project track: physical build plus
analysis. Hazard class: low (controlled bench load testing). Reader
path default: standard, with sections explicitly tagged. Named cases:
Hyatt Regency walkway collapse (1981, cross-referenced from chapter
1).}

\noindent
Chapter 1 of this volume gave the engineer the free-body diagram.
This chapter gives the engineer the static analysis it enables: how
the equilibrium conditions resolve real structures into member forces,
support reactions, and internal forces, and how the internal forces
predict the structure's response under load. Statics is the
discipline that turns the FBD's vectors into a structure's working
state.

\section{Equilibrium conditions}\pathtag{core}

A rigid body is in \engterm{static equilibrium} when the resultant
external force and the resultant external moment (about any chosen
point) both vanish. In three dimensions:
\[
\sum \vec{F}_{\text{ext}} = \vec{0}, \quad
\sum \vec{M}_{O,\,\text{ext}} = \vec{0}.
\]
The first is three scalar equations, one per coordinate direction;
the second is three more. A rigid body in three dimensions has six
independent equilibrium equations.

In two dimensions the moment equation collapses to a single scalar
(the moment about the out-of-plane axis), and the equilibrium
conditions reduce to
\[
\sum F_x = 0, \quad \sum F_y = 0, \quad \sum M_O = 0,
\]
three equations in the unknown forces and reactions.

The chosen point $O$ for the moment sum is the engineer's. Any point
works; the equations are equivalent. The choice is tactical: pick $O$
so that as many unknown forces as possible pass through $O$ and
therefore drop out of the moment equation. A standard tactic is to
take moments about a pin: the pin's two reaction components contribute
no moment about themselves, leaving only the other forces in the
moment equation. The arithmetic is shorter; the answer is the same.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-fbd-conventions.tex}

The reactions a support can supply are set by what the support
prevents. A pin prevents translation in both planar directions and
allows rotation, so it supplies two force components and no moment. A
roller prevents translation along one direction only, so it supplies a
single force normal to its running surface. A fixed (built-in) support
prevents translation in both directions and rotation, so it supplies
two force components and a moment, three reactions in all. A cable or
tie supplies a single tensile force along its own line and nothing in
compression. Reading the reactions off the supports before writing any
equation is the first act of a static analysis, and
figure~\ref{fig:vol03ch02:fbd-conventions} shows the catalogue applied
to one body: a pin at $A$, a roller at $B$, and the moment arm $L$ used
when summing moments about the pin.

A worked count fixes the determinacy idea. A planar simply-supported
beam carries a pin at one end (two reactions) and a roller at the other
(one reaction): three reactions, three equations, determinate. Add a
second roller in the middle and the count becomes four reactions
against three equations: the beam is once indeterminate, and its middle
reaction depends on how the beam bends, not on equilibrium alone.
Remove the roller from the simply-supported beam and the count drops to
two reactions against three equations: the beam is a mechanism, free to
rotate about the pin. The same arithmetic, $r$ versus three, classifies
every planar rigid body before any force is solved.

A statically determinate problem has as many independent equilibrium
equations as it has unknowns. A two-dimensional rigid body has three
equations; if its constraints produce exactly three unknowns, the
system is determinate. Fewer than three constraints and the body has
a mechanism (a degree of freedom along which it moves); more than
three and the body is statically indeterminate, and the equilibrium
equations alone cannot resolve the redundant reactions. Indeterminate
systems require the body's elastic response, which is the subject of
chapter 3 of this volume.

The count of reactions against equations classifies a body into one of
three states, and the engineer reads the state before solving. A body
with fewer independent reactions than equations is \engterm{unstable}: it
has a degree of freedom along which no support resists motion, so it is a
mechanism and equilibrium has no solution for a general load. A body with
exactly as many independent reactions as equations is \engterm{statically
determinate}: equilibrium has a unique solution. A body with more
reactions than equations is \engterm{statically indeterminate} to the
degree of the excess, and equilibrium has infinitely many solutions until
the body's stiffness picks one. The arithmetic is necessary but not
sufficient for stability, and the gap matters: three reactions whose
lines of action are concurrent, or whose lines are all parallel, leave
the body free to rotate about the concurrency point or to translate
across the parallel lines, so the body is a mechanism despite having the
determinate count. A three-legged stool with all three legs in a line is
the everyday instance. Reading the count and then checking that the
reactions are neither concurrent nor parallel is the full determinacy
test, and the case study below runs it before solving a single joint.

\begin{principle}[The three equilibrium conditions]
For a planar rigid body in static equilibrium, $\sum F_x = 0$, $\sum
F_y = 0$, and $\sum M_O = 0$ about any point $O$ in the plane. The
three equations are independent; any further moment equation about a
different point is derivable from these three and adds no new
information. The number of independent equations is fixed by the
dimensionality of the space, not by the engineer's enthusiasm for
writing more equations.
\end{principle}

Two equivalent forms of the planar conditions are often more useful
than the canonical three. The \engterm{three-moments} form:
\[
\sum M_A = 0, \quad \sum M_B = 0, \quad \sum M_C = 0,
\]
about three points $A$, $B$, $C$ that are not collinear. The
\engterm{two-moments-one-force} form:
\[
\sum F_n = 0, \quad \sum M_A = 0, \quad \sum M_B = 0,
\]
where the force sum is along an axis $n$ not perpendicular to the
line $AB$. Both forms are derivable from the canonical three. The
engineer chooses the form whose equations are simplest for the
geometry at hand.

\begin{archetype}[Balance]
The archetype of Volume III is balance. Statics is balance made
explicit: the engineer writes the resultant force and moment as
zero and solves for the reactions and member forces that enforce
the equality. The same archetype recurs in dynamics with the
addition of $m\vec{a}$ on the right-hand side, in continuum
mechanics with the integral over a stress field, and in fluid
statics with hydrostatic pressure. The mechanics of the volume is
the working out of one balance equation under successively richer
constraints on what is balancing what.
\end{archetype}

\section{Trusses: method of joints, method of sections}\pathtag{core}

A \engterm{truss} is an assembly of straight, two-force members
pinned at their ends. Each member is loaded along its axis only;
each joint is in equilibrium under the forces of the members meeting
there plus any external load applied at the joint. The truss as a
whole is in equilibrium under its support reactions plus its
external loads.

The two-force-member assumption requires that members carry no
transverse load between joints, no moment at their ends, and no
self-weight (or self-weight small compared to the design loads).
Real trusses violate the assumptions to some degree: members have
weight, joints have some rotational restraint, and members may
carry distributed loads from a deck. The truss model treats those as
small corrections to a clean idealisation; chapter 8 of this volume
returns to the corrections.

Two analytical methods resolve a truss.

\subsection{Method of joints}

Isolate each joint as a free body. The joint is a point; the forces
on it are the external loads at that joint and the axial forces of
the members meeting there. Two scalar equations of equilibrium per
joint in two dimensions ($\sum F_x = 0$, $\sum F_y = 0$); three in
three dimensions. For a truss with $j$ joints and $m$ members and
$r$ support reactions, statical determinacy in two dimensions
requires
\[
m + r = 2j,
\]
which counts unknowns (member forces and reactions) against
equations (two per joint). Trusses that satisfy this relation can be
solved joint by joint.

The procedural recipe: identify a joint at which only two unknown
member forces meet (typically a corner joint at a support). Solve
its two equations for those two unknowns. Move to an adjacent joint
where now only two unknowns remain. Continue until all members are
resolved. The procedure works when the truss is determinate and the
joints can be ordered so each one introduces only two new unknowns.
Some determinate trusses cannot be ordered that way; for those, the
method of sections fills the gap.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-method-of-joints.tex}

The bookkeeping convention that prevents most sign errors is to draw
every unknown member force as tension, pulling away from the joint
along the member's axis, as in
figure~\ref{fig:vol03ch02:method-of-joints}. The two equilibrium
equations are then written with consistent signs and solved; a member
that returns a negative number is in compression by exactly the
magnitude returned, and no diagram needs redrawing. The alternative,
guessing tension or compression member by member and flipping arrows,
multiplies the chances of a dropped sign. The same figure shows the
graphical reading: three concurrent forces in equilibrium close into a
triangle, the force polygon, whose closure is the equilibrium statement
drawn rather than computed. The polygon is the quickest independent
check on a hand calculation.

\engterm{Zero-force members} are worth spotting before any equation is
written, because they fall out of two recurring geometries. First, at a
joint where only two non-collinear members meet and no external load
acts, both members carry zero force. Second, at a joint where three
members meet, two of them collinear and the third not, and no external
load acts, the non-collinear member carries zero force. A truss
analysed by eye for its zero-force members shrinks to a smaller problem
before the first equation, and the zero-force members are not useless:
they brace compression members against buckling and carry load under a
different loading case than the one analysed. The chapter's worked Pratt
truss has one zero-force member, the central vertical $U_3L_3$, which
the numerical solver in
\texttt{code/method\_of\_joints.py} returns as $0\,\text{kN}$.

\subsection{Method of sections}

Cut the truss with an imaginary plane that passes through at most
three members in two dimensions. The cut isolates two free bodies,
each in equilibrium under the external loads and reactions on its
side plus the axial forces in the cut members. Three equilibrium
equations on each side, three unknowns (the three cut member
forces), and the cut members' forces fall out directly.

The method of sections is the right tool when the engineer wants the
force in a specific interior member without solving every joint
between it and the support. A typical case is the analysis of a
deep truss in which the chord forces grow with the bending moment;
a single section near midspan gives the maximum chord force in one
calculation, where the method of joints would require working
inward from a support.

The three cut-member forces can each be isolated by choosing the
moment point well. To find a chord force, take moments about the joint
where the other two cut members meet, which removes them and leaves one
equation in the chord force. To find a diagonal force, the two chords
are usually parallel; sum forces perpendicular to the chords and the
chord contributions vanish, leaving the diagonal alone. The method of
sections rewards the same tactical thinking as the choice of moment
point in a global analysis: choose the equation that isolates the
unknown wanted.

\subsection{Assumptions and their corrections}

The truss model assumes pin joints, axial-only loading, and
no member weight. The assumptions are corrections from real
behaviour:

\begin{itemize}[leftmargin=*]
  \item Real joints are bolted, welded, or gusseted; they have
    rotational stiffness. A truss analysed as pin-jointed and built
    with rigid joints will distribute load to all members, not just
    the geometrically determinate ones. The corrected analysis is
    the rigid-frame analysis of chapter 8 of this volume.
  \item Real members have weight, distributed along their length.
    The standard correction lumps half the member weight at each
    end joint. The correction is accurate for slender members and
    becomes inaccurate when the member weight is comparable to the
    applied load.
  \item Real deck and roof loads are distributed, not point loads
    at joints. The standard correction is to lump distributed load
    into equivalent point loads at the joints it acts between, by
    the tributary-area rule. The correction is exact when the deck
    is simply supported between joints (no bending stiffness
    transfer); approximate otherwise.
\end{itemize}

A truss whose loads do not act at the joints, or whose members
carry transverse load between joints, cannot be analysed as a truss.
The structure is a \engterm{frame}, and frames are the subject of
the next section.

\begin{mastery}
The method of joints, written as a single linear system, is the matrix
method that every truss-analysis program runs. Assign each joint two
equilibrium equations and stack them into a column vector of length $2j$.
Each unknown, the $m$ member forces and the $r$ reactions, gets a column.
A member running from joint $a$ to joint $b$ has unit vector
$\hat{u} = (p_b - p_a)/\lVert p_b - p_a \rVert$; carrying its force as
tension, it contributes $+\hat{u}$ to joint $a$'s two rows and $-\hat{u}$
to joint $b$'s. A reaction contributes its own unit direction to the one
joint it acts on. Collecting the contributions builds the equilibrium
matrix $\mathbf{A}$ of size $2j \times (m + r)$, and the system is
\[
\mathbf{A}\, \vec{x} = -\vec{p},
\]
where $\vec{x}$ holds the member forces and reactions and $\vec{p}$ holds
the applied joint loads. For a determinate truss, $m + r = 2j$,
$\mathbf{A}$ is square, and the single solve $\vec{x} = -\mathbf{A}^{-1}
\vec{p}$ returns every member force and reaction at once, with a positive
entry meaning tension and a negative meaning compression, the same sign
convention the hand method uses.

The matrix carries diagnostics the hand method hides. If $\mathbf{A}$ is
singular, the truss is a mechanism (its geometry admits a motion that no
member resists), and the singularity flags the defect that the
determinacy count $m + r = 2j$ alone cannot: a truss can satisfy the
count and still be a mechanism if its members are badly arranged, for
instance four members forming a parallelogram that shears. The condition
number of $\mathbf{A}$ measures how close the truss sits to such a
mechanism; a condition number that spikes as a design is tweaked warns
that the truss is approaching a geometry in which small load changes
produce large member-force changes, the numerical signature of a
near-mechanism. The solver in \texttt{code/method\_of\_joints.py}
assembles exactly this $\mathbf{A}$, checks the count, reports the
condition number, and solves; the chapter's Pratt and Howe results are
its output. The same assembly, with member stiffnesses $EA/L$ replacing
unit equilibrium and displacements replacing forces as the unknowns, is
the direct stiffness method of chapter 8, which carries the analysis from
determinate trusses to indeterminate frames by adding the members'
elastic response.
\end{mastery}

\section{Frames and machines}\pathtag{standard}

A \engterm{frame} is an assembly of members at least one of which is
loaded between its endpoints. A \engterm{machine} is an assembly of
members designed to transmit or modify forces, typically with
moving parts. Both differ from a truss in that members may carry
bending moments and transverse forces, not only axial forces.

The analysis tactic for a frame or machine is to disassemble it
into its component members and apply the equilibrium conditions to
each member separately. Each member's FBD shows the external loads
on that member plus the forces transmitted by its connections to
other members. By Newton's third law, the force one member exerts
on another at a pin is equal and opposite to the force the other
exerts back. Bookkeeping is the discipline: if member $AB$ exerts
$\vec{F}_{AB \to BC}$ on member $BC$ at pin $B$, then $BC$ exerts
$-\vec{F}_{AB \to BC}$ on $AB$ at the same pin. Crossing the sign on
one of the two and not the other is the surest way to a wrong
answer.

The unknowns in a frame analysis are the support reactions and the
pin forces at each interior connection. A frame with $n$ members
and $p$ interior pin connections in two dimensions has, as
equations, three per member ($3n$) and, as unknowns, the support
reactions plus two pin-force components at each interior pin ($r +
2p$). Statical determinacy requires $3n = r + 2p$. Frames that
violate the relation are indeterminate and require the elastic
analysis of chapter 8.

The two-force-member shortcut speeds frame analysis substantially.
Any member of a frame that carries no transverse load between its
endpoints and no moment at its endpoints can be analysed as a
two-force member; the direction of the force in such a member is
along the member's axis. Identifying two-force members at the
start of an analysis reduces the unknown count and lets the
engineer write fewer equations.

\engterm{Machines} differ from frames in that the engineer
typically wants to compute the relationship between an input force
and an output force, given the geometry. The classical examples are
the lever, the wedge, the pulley, the screw, the inclined plane,
and the toggle. Each is analysed by the same equilibrium discipline
as a frame, with the input and output forces named as unknowns and
the geometry's mechanical advantage emerging from the solution.
Coulomb's 1781 prize memoir on simple machines remains the
historical reference for the empirical corrections from idealised
to real machines, the largest of which is rope-and-pulley friction
\cite{hist:coulomb1781}.

The \engterm{mechanical advantage} of a machine is the ratio of output
force to input force, and statics reads it straight off the moment
equation. For a lever, equilibrium about the fulcrum gives $F_{\text{out}}
\, a_{\text{out}} = F_{\text{in}}\, a_{\text{in}}$, so the advantage is the
ratio of arms $a_{\text{in}}/a_{\text{out}}$: a crowbar with a long handle
and a short toe multiplies a hand force into a prying force many times
larger. The same balance governs the others. A pulley block of $n$
supporting rope falls carries each fall at the load tension divided by
$n$, an ideal advantage of $n$. A screw of pitch $p$ and mean thread
radius $r$ converts a turning torque into an axial thrust with ideal
advantage $2\pi r/p$, which is large because the pitch is small. The
common thread is that an ideal machine trades force for distance with the
product conserved, so the work in equals the work out, and the mechanical
advantage is the geometric ratio the moment equation returns. Real
machines lose part of the trade to friction, the loss Coulomb's memoir
quantified, and the band brake worked later in this chapter shows the
opposite use of the same friction, turned from a loss into the machine's
whole purpose.

\section{Distributed loads}\pathtag{standard}

A distributed load is a force per unit length (a line load), per
unit area (a surface load), or per unit volume (a body load). Wind
on a building, water on a dam, snow on a roof, weight of a beam,
gravity on a parked car: all are distributed.

For the equilibrium of a rigid body, a distributed load can be
replaced by an equivalent point load whose magnitude equals the
total load (the integral of the distribution over its domain) and
whose line of action passes through the load's centroid. The
equivalent point load gives the correct resultant force and the
correct resultant moment about any point; it does not give the
correct stress distribution within the body, which depends on
where the load is actually applied and is the subject of chapter 3
of this volume.

For a line load $w(x)$ over a span $L$, the equivalent point load
is
\[
W = \int_0^L w(x) \,dx,
\]
acting at the centroid of the loading curve,
\[
\bar{x} = \frac{1}{W} \int_0^L x \, w(x) \,dx.
\]
Two common cases: a uniform line load $w(x) = w_0$ has $W = w_0 L$
at midspan; a triangular line load growing linearly from zero to
$w_{\max}$ has $W = \tfrac{1}{2} w_{\max} L$ at two-thirds span
from the zero end.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-distributed-loads.tex}

Figure~\ref{fig:vol03ch02:distributed-loads} collects the three cases
the working engineer meets daily. Any piecewise-linear distribution is
a sum of rectangles and triangles; replace each piece by its own
resultant at its own centroid, then combine the pieces with the same
area-weighting that combines composite centroids. The trapezoid in the
figure is the canonical instance: split it into a rectangle of height
$w_1$ and a triangle of peak $w_2 - w_1$, locate each resultant, and
sum. The decomposition turns an integral into mental arithmetic.

The integral can be done geometrically. The total load is the
area under the loading diagram; the centroid is the centroid of
that area. For piecewise-linear distributions (rectangles,
triangles, trapezoids), the integral is a sum of areas and the
centroid a weighted average of the constituent centroids, both
done in one's head once the engineer has internalised the
geometry of the standard shapes.

Surface and volume loads use the same approach with the
appropriate two- or three-dimensional integrals. Hydrostatic
pressure on a vertical wall, for instance, is a triangular line
load when expressed per unit horizontal length of wall, with
resultant force per unit length $\rho g h^2 / 2$ acting at one-third
the water depth from the bottom.

\section{Centroids and moments of area}\pathtag{standard}

The centroid of a planar region is the point through which a
parallel uniform line load over the region would act as an
equivalent point load. The centroid of a homogeneous rigid body is
the point at which its weight may be lumped for moment calculations.
For a homogeneous body, the centroid coincides with the centre of
mass; for a body with varying density, the two differ.

For a planar region of area $A$:
\[
\bar{x} = \frac{1}{A} \int_A x \, dA, \quad
\bar{y} = \frac{1}{A} \int_A y \, dA.
\]
For composite regions, the centroid is the area-weighted average
of constituent centroids,
\[
\bar{x}_{\text{total}} = \frac{\sum_i A_i \bar{x}_i}{\sum_i A_i},
\]
and similarly for $\bar{y}$. Hollow regions are handled by
subtracting the missing area with a negative sign.

The \engterm{first moment of area} about an axis is $\int_A x \,dA$
(about the $y$-axis) or $\int_A y \,dA$ (about the $x$-axis). The
centroid is the point through which the first moment vanishes: by
definition, $\int_A (x - \bar{x}) dA = 0$.

The \engterm{second moment of area}, also called the area moment of
inertia, is
\[
I_x = \int_A y^2 \,dA, \quad I_y = \int_A x^2 \,dA, \quad
J_O = \int_A r^2 \,dA = I_x + I_y,
\]
the last being the polar moment about the origin. The second moment
of area appears in the bending-stress formula $\sigma = My/I$ (chapter
8 of this volume), in the buckling load $P_{cr} = \pi^2 EI / L^2$
(also chapter 8), and in the torsional rigidity of a circular shaft.
The dimension is $[\text{length}]^4$; an area of one square metre
distributed at radius one metre about the axis has second moment of
$1\,\si{\meter\tothe{4}}$.

\engterm{Parallel-axis theorem}: the second moment about an axis
parallel to a centroidal axis is
\[
I = I_C + A d^2,
\]
where $I_C$ is the second moment about the centroidal axis and $d$
the perpendicular distance between the axes. The theorem lets the
engineer compute second moments of composite sections (I-beams,
T-sections, channels) by summing the contributions of constituent
rectangles, each shifted to the composite's centroid by the
parallel-axis term. The largest contribution comes from the area
farthest from the centroidal axis, which is why I-beams and
hollow tubes are stiffer per unit mass than solid cross-sections.

Standard tables of centroids and second moments of area for common
shapes appear in mechanics textbooks and engineering handbooks
\cite[appendices]{text:meriam-kraige2018}. The working engineer
internalises a handful of the most common (rectangle, triangle,
circle, semicircle, I-section) and computes the rest as needed. A
compact table of the five most common shapes ships with this chapter
as \texttt{data/section\_table.csv}, and the composite-section routine
\texttt{code/section\_properties.py} applies the parallel-axis theorem
automatically to any stack of rectangles.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-centroid-composite.tex}

Figure~\ref{fig:vol03ch02:centroid-composite} shows the parallel-axis
theorem at work on an I-section. The section's stiffness is dominated by
the two flanges, each carrying a transfer term $A_i d_i^2$ that grows
with the square of its distance from the centroidal axis. The web,
sitting on the centroidal axis, contributes its small $bh^3/12$ and no
transfer term. This is why an I-beam is the standard bending shape: it
spends material where the transfer term pays the most, at the extreme
fibres, and leaves only a thin web to carry the shear and hold the
flanges apart. The numerical routine returns
$I_x \approx 1.04 \times 10^{-4}\,\si{\meter\tothe{4}}$ for the
$200/20$ flange, $200/10$ web section, which the I-section exercise
checks against a hand calculation.

\begin{estimation}
\textbf{Estimate the second moment of area of a $300 \,\si{\milli
\meter}$ deep rectangular wooden beam, $50\,\si{\milli\meter}$ wide,
about its centroidal axis perpendicular to the depth.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The second moment of a rectangular section about its centroidal
axis is $I = bh^3 / 12$, where $b$ is the width perpendicular to
the load and $h$ is the depth (height) parallel to the load. The
key observation is the cube on $h$: doubling the beam's depth
multiplies the second moment by eight, while doubling the width
multiplies it by only two.

For our beam, $b = 50\,\si{\milli\meter} = 0.05\,\si{\meter}$,
$h = 300\,\si{\milli\meter} = 0.3\,\si{\meter}$:
\[
I = \frac{0.05 \times 0.3^3}{12} =
\frac{0.05 \times 0.027}{12} =
\frac{1.35 \times 10^{-3}}{12} \approx
1.13 \times 10^{-4} \,\si{\meter\tothe{4}}.
\]
Estimate: $\sim 10^{-4} \,\si{\meter\tothe{4}}$.

The postmortem: orient the beam with its $300\,\si{\milli\meter}$
dimension vertical, perpendicular to the load. If the beam were
oriented the other way (width vertical), the second moment would be
$I' = 0.3 \times 0.05^3 / 12 \approx 3.1 \times 10^{-6}\,\si{\meter
\tothe{4}}$, a factor of $36$ smaller. Carpenters who set a joist on
its narrow edge rather than its flat are exploiting the cube: a
joist on edge resists deflection thirty-six times better than the
same joist laid flat. The same fact explains why a sheet of paper
held horizontally between fingertips collapses under its own weight
but held by an edge curl supports a small object.
\end{estimation}

\section{Internal forces: shear, moment, axial}\pathtag{core}

A rigid body in external equilibrium has internal forces at every
cross-section. The internal forces transmit the external loads from
one part of the body to another. Cut the body at any plane, and the
two halves are themselves rigid bodies in equilibrium: each carries
the external loads on its side plus the internal forces and
moments transmitted across the cut. The internal forces are not
arbitrary; they are determined by the equilibrium of either half.

For a planar straight member (a beam), the internal force at a
cross-section resolves into three components:

\begin{itemize}[leftmargin=*]
  \item \engterm{Axial force} $N$, along the beam's axis. Positive
    in tension by convention.
  \item \engterm{Shear force} $V$, perpendicular to the beam's
    axis. Sign convention varies by textbook; the convention used
    in this work is that $V$ is positive when the right-hand side
    of the cut is pushed downward (or equivalently, when the
    left-hand side is pushed upward).
  \item \engterm{Bending moment} $M$, about the out-of-plane axis.
    Positive when the beam is bent concave-up (sagging), negative
    when concave-down (hogging).
\end{itemize}

The three internal-force components are functions of position along
the beam: $N(x)$, $V(x)$, $M(x)$. The shear and moment diagrams
plot $V(x)$ and $M(x)$ versus $x$ and are the central deliverable of
beam analysis. The points of maximum $|V|$ and maximum $|M|$ are the
critical sections for shear failure and bending failure
respectively.

Two differential relations connect the distributed load $w(x)$ (per
unit length, positive downward), the shear, and the moment:
\[
\frac{dV}{dx} = -w(x), \quad \frac{dM}{dx} = V(x).
\]
The first follows from vertical equilibrium of an infinitesimal
element; the second from moment equilibrium of the same element.
Together they imply
\[
\frac{d^2 M}{dx^2} = -w(x),
\]
the second-order equation that, with boundary conditions, recovers
the moment distribution from the load.

The differential relations give the practical engineer three
geometric rules for sketching shear and moment diagrams without
integration:

\begin{itemize}[leftmargin=*]
  \item The slope of the shear diagram equals the negative of the
    load intensity. Concentrated point loads produce a jump in
    shear equal to the load; distributed loads produce a sloped
    shear curve.
  \item The slope of the moment diagram equals the shear. The
    moment is maximum (or minimum) at points where the shear
    crosses zero.
  \item Concentrated applied moments produce a jump in the moment
    diagram equal to the applied moment. Concentrated point loads
    do not produce a jump in the moment diagram; they produce a
    kink (a change in slope of $M$ equal to the size of the load).
\end{itemize}

The rules let the engineer draw $V(x)$ and $M(x)$ for any
simply-supported beam with point and distributed loads in less
than a minute, without integrating. The maximum bending moment of a
simply-supported beam with span $L$ and centred point load $P$ is
$M_{\max} = PL/4$ at midspan; with uniform line load $w_0$ it is
$M_{\max} = w_0 L^2 / 8$ at midspan. The two cases come up
constantly and are worth memorising.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-shear-moment.tex}

Figure~\ref{fig:vol03ch02:shear-moment} draws the central-point-load
case in full: the rectangular shear diagram, the triangular moment
diagram, and the peak $PL/4$ located exactly where the shear crosses
zero. The accompanying routine
\texttt{code/shear\_moment.py} integrates $-w$ to shear and shear to
moment numerically for any combination of point loads, distributed
loads, and applied couples, and reproduces both memorised peaks to
within the discretisation error.

\begin{mastery}
The differential relations $dV/dx = -w$ and $dM/dx = V$ are the beam's
form of a conservation law, and they hold piecewise between
discontinuities. At a concentrated point load the shear jumps and the
relation $dV/dx = -w$ holds in the distributional sense, with the point
load a Dirac delta in $w(x)$; integrating across the load recovers the
finite jump. At a concentrated applied couple the moment jumps while the
shear is continuous, the couple appearing as a delta in the second
derivative. Treating point loads and couples as singular distributions
unifies the discrete and continuous cases under one pair of equations,
which is the form a finite-element formulation generalises: the element
load vector is the integral of $w(x)$ against the element shape
functions, and a point load is the special case where $w$ is a delta and
the integral picks out the shape-function value at the load. The reader
who carries the relations forward to Volume V's continuum mechanics will
recognise $dV/dx = -w$ as the one-dimensional restriction of the
equilibrium of stresses, $\nabla \cdot \boldsymbol{\sigma} +
\vec{b} = \vec{0}$, with the shear playing the role of the stress
resultant and $w$ the body force per unit length.
\end{mastery}

\section{Worked examples}\pathtag{standard}

\subsection{A Pratt truss under uniform deck load}

A simply-supported Pratt truss spans $L = 12\,\si{\meter}$, has
height $h = 2\,\si{\meter}$, and consists of six bays each
$2\,\si{\meter}$ wide. Deck loading delivers a downward joint
load of $P = 20\,\text{kN}$ to each of the five interior bottom
chord joints. Find the support reactions, the maximum chord force,
and the force in the diagonal in the third bay from the left
support.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-pratt-truss.tex}

\textbf{Reactions.} Symmetry gives equal vertical reactions at the
two supports. The total downward load is $5P = 100\,\text{kN}$,
so each reaction is $V_A = V_B = 50\,\text{kN}$ upward.

\textbf{Maximum chord force.} The bending moment of the truss as a
whole, computed as for a simply-supported beam with the joint
loads as concentrated point loads, is maximum at midspan. With
five equally spaced loads at $2,\,4,\,6,\,8,\,10\,\si{\meter}$ from
the left support, the moment at midspan ($x = 6\,\si{\meter}$) is
\[
M_{\max} = 50 \times 6 - 20 \times 4 - 20 \times 2 =
300 - 80 - 40 = 180\,\text{kN}\cdot\text{m}.
\]
The chord forces at midspan resist this moment as a couple with
arm equal to the truss depth $h$:
\[
F_{\text{chord}} = \frac{M_{\max}}{h} = \frac{180}{2} =
90\,\text{kN}.
\]
The top chord is in compression, the bottom chord in tension.

\textbf{Diagonal in the third bay.} A vertical section through
the third bay, just to the right of the second interior bottom
joint, cuts the top chord, the bottom chord, and the diagonal. The
left free body has the left reaction $V_A = 50\,\text{kN}$, the
two joint loads to its left ($20\,\text{kN}$ each), and three
unknown cut member forces.

Vertical equilibrium of the left free body:
\[
50 - 20 - 20 - F_d \sin\theta = 0,
\]
where $\theta = \arctan(h / \text{bay width}) = \arctan(2/2) =
45^{\circ}$. Solving, $F_d = 10 / \sin 45^{\circ} = 10\sqrt{2}
\approx 14.1\,\text{kN}$. The sign is positive (we assumed the
diagonal carries the left free body upward, requiring tension in
the diagonal toward the upper joint); a check of the geometry
confirms the diagonal is in tension.

A full analysis would proceed joint by joint to verify; the
section's value is that it gives one member's force in one
calculation. The numerical solver
\texttt{code/method\_of\_joints.py} resolves all twenty-one members at
once and confirms the section result: it returns the diagonal $U_2L_3$
as $14.14\,\text{kN}$ in tension, the maximum compressive chord force as
$90\,\text{kN}$ in the two central top-chord members, and the central
vertical $U_3L_3$ as a zero-force member. The full member-force table is
shipped as \texttt{data/pratt\_truss\_forces.csv}.

\textbf{Maximum chord force by section.} The same section gives the
chord forces directly. Take moments of the left free body about the top
joint $U_2$, where the diagonal and top chord both pass and so drop out.
Only the bottom chord $L_2L_3$ and the known loads remain. With $U_2$ at
$x = 4\,\si{\meter}$ and height $2\,\si{\meter}$,
\[
F_{L_2L_3} \cdot h = V_A \cdot 4 - 20 \cdot 2,\quad
F_{L_2L_3} = \frac{50 \cdot 4 - 20 \cdot 2}{2} = 80\,\text{kN},
\]
in tension, matching the solver. Taking moments about the bottom joint
$L_3$ instead isolates the top chord $U_2U_3$ and returns
$90\,\text{kN}$ in compression, the maximum chord force, in one line.

\subsection{Case study: a Howe roof truss carried end to end}

The Pratt example demonstrated each method on its own member. A second
truss, carried from raw geometry to a design verdict without skipping a
step, shows how the pieces assemble into a complete analysis, the form an
engineer's calculation package actually takes. We analyse a Howe roof
truss: the gable-profile truss that carries pitched-roof loads in timber
and light steel construction across much of the world.

\textbf{Geometry and loads.} The truss spans $L = 16\,\si{\meter}$
between a pin support at the left and a roller at the right, divided into
eight bottom-chord bays of $2\,\si{\meter}$. It rises to a ridge
$4\,\si{\meter}$ above the supports, so each rafter climbs
$4\,\si{\meter}$ over a horizontal run of $8\,\si{\meter}$, a slope of
$\arctan(4/8) = \arctan(0.5) \approx 26.57^{\circ}$. The bottom chord
runs level along the eight joints $L_0$ through $L_8$; the top chord
follows the gable through the seven interior apex joints $T_1$ through
$T_7$, with $T_4$ at the ridge. Verticals drop from each top-chord joint
to the bottom chord beneath it; the Howe diagonals slope down toward
midspan. Purlins deliver an equal downward load $P = 10\,\text{kN}$ to
each of the seven top-chord joints, the lumped tributary of the roof
sheathing, snow, and self-weight.
Figure~\ref{fig:vol03ch02:howe-roof-truss} fixes the geometry.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-howe-roof-truss.tex}

\textbf{Determinacy check, first.} The truss has $j = 16$ joints (nine
bottom, seven top), $m = 29$ members, and $r = 3$ reactions (two at the
pin, one at the roller). The count $m + r = 29 + 3 = 32$ equals
$2j = 32$, so the truss is statically determinate and the method of
joints will resolve it. Running the count before any force is solved
catches a mis-drawn truss while the error is still cheap: a missing
diagonal would have given $m + r = 31 < 32$, the signature of a
mechanism, and an extra member would have given $33 > 32$, the signature
of internal indeterminacy that equilibrium alone cannot resolve.

\textbf{Support reactions.} The loading is symmetric about the ridge, and
the supports sit at equal heights, so the two vertical reactions are
equal. The total downward load is $7P = 70\,\text{kN}$, so
\[
R_A = R_B = \tfrac{1}{2}(7P) = 35\,\text{kN} \text{ upward}.
\]
The pin also supplies a horizontal reaction, which the horizontal
equilibrium $\sum F_x = 0$ sets to zero because no external load has a
horizontal component. The roller carries no horizontal force by
construction.

\textbf{Zero-force members, spotted by eye.} Joint $L_1$ carries the
end of the bottom chord, the first diagonal, and the vertical $T_1L_1$.
The bottom chord and the diagonal are the two members whose forces have
to balance the horizontal direction at $L_1$; the vertical $T_1L_1$ is
the odd member out, and with no external load at $L_1$ its force is
zero. The mirror joint $L_7$ gives the second zero-force member $T_7L_7$
by the same reasoning. Both verticals carry no force under this
symmetric gravity case; they earn their place by bracing the rafters
against buckling and by carrying load under the unbalanced snow cases the
roof code also requires.

\textbf{Member forces by joints.} Begin at the pin $L_0$, where only two
members meet: the bottom chord $L_0L_1$ (horizontal) and the end rafter
$L_0T_1$ (climbing at $26.57^{\circ}$). The two equilibrium equations
resolve both. Vertical equilibrium balances the reaction against the
rafter's vertical component:
\[
R_A + F_{L_0T_1}\sin 26.57^{\circ} = 0, \quad
F_{L_0T_1} = \frac{-35}{0.4472} = -78.26\,\text{kN},
\]
negative, so the end rafter is in compression at $78.26\,\text{kN}$.
Horizontal equilibrium then gives the bottom chord:
\[
F_{L_0L_1} + F_{L_0T_1}\cos 26.57^{\circ} = 0, \quad
F_{L_0L_1} = -(-78.26)(0.8944) = 70.0\,\text{kN},
\]
positive, the bottom chord in tension at $70\,\text{kN}$. Moving joint by
joint inward, each new joint introducing at most two unknowns, resolves
the rest. The bottom chord falls from $70\,\text{kN}$ at the supports to
$50\,\text{kN}$ in the two panels flanking the ridge; the rafters fall
from $78.26\,\text{kN}$ at the ends to $44.72\,\text{kN}$ at the ridge;
the tension verticals run from zero at the ends through $5$, $10$, and
$30\,\text{kN}$ at the ridge vertical $T_4L_4$; and the compression
diagonals run from $11.18\,\text{kN}$ to $18.03\,\text{kN}$ toward
midspan.
Figure~\ref{fig:vol03ch02:howe-force-map} maps the full result, tension
in green, compression in red, the two zero-force members dashed.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-howe-force-map.tex}

\textbf{Spot-check by sections.} A single calculation confirms the
joint-by-joint march on the most-loaded interior panel. Cut a vertical
section between $L_3$ and $L_4$; it crosses the top chord $T_3T_4$, the
diagonal $T_3L_4$, and the bottom chord $L_3L_4$. Take the left free
body, which carries the reaction $R_A = 35\,\text{kN}$ at $L_0$ and the
three purlin loads of $10\,\text{kN}$ at $T_1$, $T_2$, $T_3$. To isolate
the bottom chord, sum moments about $T_3$ at $(6, 3)$, where the top
chord and diagonal both pass and so drop out. The bottom chord, acting
horizontally at height zero, has moment arm $3\,\si{\meter}$ about
$T_3$:
\[
F_{L_3L_4}\cdot 3 = R_A \cdot 6 - 10 \cdot 4 - 10 \cdot 2
= 210 - 40 - 20 = 150,\quad
F_{L_3L_4} = 50\,\text{kN},
\]
in tension, matching the joint march exactly. Summing moments instead
about $L_4$ at $(8, 0)$, where the bottom chord and diagonal both pass,
isolates the top chord and returns $44.72\,\text{kN}$ in compression,
again matching. The section gives two of the panel's three member forces
in two lines, the independent check that the joint-by-joint bookkeeping
did not drift.

\textbf{Governing members and the verdict.} Two findings set the design.
The end rafters carry $78.26\,\text{kN}$ in compression over an unbraced
length of one rafter panel, $\sqrt{2^2 + 1^2} = 2.24\,\si{\meter}$; they
are the buckling-critical members and size the rafter section through
$P_{cr} = \pi^2 EI / L_e^2$ (chapter 8 of this volume). The four
support-adjacent bottom-chord panels carry $70\,\text{kN}$ in tension;
they are net-section-critical at their bolted splices and size the bottom
chord through the net-section and block-shear checks of this chapter's
failure section. The interior diagonals and verticals carry well under
half the governing forces and are sized by minimum-member rules and
buckling of the compression diagonals, not by strength. The verdict: a
designer who sizes the end rafters for $78\,\text{kN}$ compression with a
buckling check and the support bottom chord for $70\,\text{kN}$ tension
with a connection check has caught the two members that govern, and the
remaining twenty-seven follow. The numerical solver of
\texttt{code/method\_of\_joints.py} reproduces every force in this
analysis when given the Howe connectivity; the analysis above is what the
solver automates, carried by hand so the reader sees where each number
comes from.

\subsection{Case study: a parabolic suspension cable carried end to end}

The two trusses resolved into discrete members, each carrying a single
axial force. A cable is the opposite limit: a continuous member that can
carry only tension and reshapes itself until the tension is the only
internal force it needs. We carry a suspension cable from its loading to
its tower force and its anchor pull, the analysis that sizes the main cable
of a footbridge or the catenary of a small suspension span.

\textbf{Geometry and loads.} A cable hangs between two towers of equal
height a span $L = 60\,\si{\meter}$ apart and sags $h = 6\,\si{\meter}$ at
midspan. A stiffening deck hangs from the cable through closely spaced
vertical hangers and delivers a load $w = 12\,\text{kN}/\text{m}$ measured
per unit of \emph{horizontal} length. The hangers being close together, the
cable load is uniform in the horizontal coordinate, which is the condition
that makes the cable a parabola rather than the catenary it would take under
its own distributed weight along the arc.
Figure~\ref{fig:vol03ch02:parabolic-cable} fixes the geometry.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-parabolic-cable.tex}

\textbf{Horizontal tension, first.} The horizontal component of the cable
tension, call it $H$, is constant along the whole cable, because no
horizontal load acts on the cable between the towers and horizontal
equilibrium of any cable segment carries $H$ through unchanged. Isolate the
half-cable from the midspan low point to the left tower. At the low point
the cable is horizontal, so it pulls the half-cable inward with the force
$H$ alone. The deck load on that half is $w(L/2) = 12 \cdot 30 =
360\,\text{kN}$, acting down at its own centroid, a quarter-span
$L/4 = 15\,\si{\meter}$ from the tower. Summing moments about the tower top
removes the tower reaction and leaves $H$ against the deck load:
\[
H \cdot h = w\,\frac{L}{2}\cdot \frac{L}{4}, \quad
H = \frac{w L^2}{8 h} = \frac{12 \cdot 60^2}{8 \cdot 6}
= \frac{43200}{48} = 900\,\text{kN}.
\]
The horizontal tension is the same formula as the midspan moment of a
simply-supported beam divided by the sag, $H = (wL^2/8)/h$, the sag playing
the role the truss depth played in the chord-force formula. A flatter cable,
smaller $h$, carries a larger horizontal tension for the same load, the
trade the next paragraph makes explicit.

\textbf{Tower tension, the maximum.} The cable tension at any point is the
resultant of the constant horizontal $H$ and the vertical shear the cable
has picked up from the low point to that point. The vertical force the
half-cable hands to the tower is the full half-deck load,
\[
V = w\,\frac{L}{2} = 12 \cdot 30 = 360\,\text{kN},
\]
so the cable tension is largest at the tower, where both components are
present:
\[
T_{\max} = \sqrt{H^2 + V^2} = \sqrt{900^2 + 360^2}
= \sqrt{810000 + 129600} = \sqrt{939600} \approx 969\,\text{kN}.
\]
At the midspan low point the vertical component is zero and the tension
falls to its minimum, $T_{\min} = H = 900\,\text{kN}$. The cable is sized
for the tower tension of $969\,\text{kN}$, and the slope of the cable at the
tower, $\arctan(V/H) = \arctan(360/900) = 21.8^{\circ}$, is the angle the
saddle and the back-stay must turn.

\textbf{Support reactions and the anchor pull.} Each tower carries the
vertical reaction $V = 360\,\text{kN}$ down into its foundation, the
half-deck load delivered as a compression in the tower. The horizontal
tension $H = 900\,\text{kN}$ does not vanish at the tower: the main cable
continues over a saddle to a back-stay that runs down to a buried anchor
block, and the anchor must resist the horizontal pull of $900\,\text{kN}$
(more, once the back-stay's own angle is accounted for). An anchor sized for
the deck weight alone, $360\,\text{kN}$, would be under-specified by a
factor of two and a half, the same geometry surprise the bracket crane
showed, here at the scale of a buried concrete block.

\textbf{The sag-versus-tension trade, and the verdict.} The horizontal
tension $H = wL^2/(8h)$ runs inversely with the sag. Halving the sag to
$3\,\si{\meter}$ would double $H$ to $1800\,\text{kN}$ and push the tower
tension above $1830\,\text{kN}$, nearly doubling the cable and anchor sizes;
doubling the sag to $12\,\si{\meter}$ would halve $H$ at the cost of taller
towers and a deeper structure. The designer's free variable is the sag, and
it trades cable and anchor force against tower height and clearance. The
verdict for this span: a sag near a tenth of the span ($h/L = 0.1$) sets the
tower tension near $970\,\text{kN}$ and the anchor pull near $900\,\text{kN}$,
both governed by the horizontal tension the sag controls, and a designer who
sizes the cable for the tower tension and the anchor for the horizontal pull
has caught the two forces that govern a suspension structure.

\begin{mastery}
The cable in the case study took a parabola because its load was uniform in
the horizontal coordinate. A cable hanging under its own weight, with no
deck, carries a load uniform along its \emph{arc length} instead, and the
shape it takes is the catenary, not the parabola. The distinction follows
from the same horizontal-equilibrium argument with a different load measure.
Let $H$ be the constant horizontal tension and let $s$ measure arc length
from the low point. For a self-weight $w_s$ per unit arc length, vertical
equilibrium of the segment from the low point gives the cable slope as
$dy/dx = w_s s / H$, and combining with $ds = \sqrt{1 + (dy/dx)^2}\,dx$
integrates to
\[
y(x) = \frac{H}{w_s}\Big(\cosh\frac{w_s x}{H} - 1\Big),
\]
the catenary, named from the Latin \emph{catena} for chain because a free
chain hangs in exactly this curve. The parabola of the deck-loaded cable is
$y(x) = (w x^2)/(2H)$ with $w$ per unit horizontal length, the leading term
of the catenary's series expansion $\cosh u - 1 = u^2/2 + u^4/24 + \cdots$.
The two curves agree closely when the sag is small relative to the span, the
shallow-cable regime where $\cosh$ is near its parabolic approximation, and
diverge for deep cables where the higher terms matter. The engineering
reading: a suspension bridge whose deck weight dwarfs the cable weight is
analysed as a parabola with negligible error; a transmission line or a free
chain, carrying only its own weight, is a catenary, and using the parabola
for a deep transmission span understates the tension at the towers. The
chapter's case study sits firmly in the parabolic regime, with the deck load
$12\,\text{kN}/\text{m}$ many times the cable's own weight per metre.
\end{mastery}

\subsection{A Warren truss interior diagonal by one section}

The Pratt example took a section to find a single diagonal; the same cut
finds a Warren-truss diagonal, and the Warren's alternating geometry shows
how the method of sections isolates one member without touching the joints
around it. A Warren truss spans $L = 12\,\si{\meter}$ in six bottom-chord
panels of $2\,\si{\meter}$, has a constant depth $h = 2\,\si{\meter}$, and
carries equal downward loads $P = 24\,\text{kN}$ at the four interior
bottom-chord joints. The diagonals alternate in slope, up-and-down across
the span, with no verticals, the defining Warren pattern.

\textbf{Reactions.} The total load is $4P = 96\,\text{kN}$, symmetric, so
each support reaction is $R = 48\,\text{kN}$ upward.

\textbf{The diagonal by a vertical cut.} Cut a vertical section through the
second panel from the left, crossing the top chord, one diagonal, and the
bottom chord. The diagonal climbs at $\arctan(h/\text{panel}) =
\arctan(2/2) = 45^{\circ}$. Take the left free body, which carries the
reaction $R = 48\,\text{kN}$ at the left support and the single interior
load $P = 24\,\text{kN}$ at the first bottom joint, $2\,\si{\meter}$ in. The
two chords run horizontal, so summing forces in the vertical direction
removes both chords and isolates the diagonal in one equation:
\[
48 - 24 - F_d \sin 45^{\circ} = 0, \quad
F_d = \frac{24}{0.7071} = 33.9\,\text{kN}.
\]
The sign is positive against the assumed tension, so this diagonal is in
tension at $33.9\,\text{kN}$. The neighbouring diagonal, sloping the other
way, carries the same magnitude in compression by the alternating geometry,
which is the practical reason a Warren truss alternates tension and
compression diagonals along its length and lets the designer make the
tension diagonals slender.

\textbf{A chord by a moment.} The same section gives a chord force from one
moment equation. Sum moments of the left free body about the top joint where
the diagonal and top chord meet, $4\,\si{\meter}$ from the support at height
$2\,\si{\meter}$; both cut members above drop out, leaving the bottom chord:
\[
F_{\text{bottom}} \cdot 2 = 48 \cdot 4 - 24 \cdot 2 = 192 - 48 = 144, \quad
F_{\text{bottom}} = 72\,\text{kN},
\]
in tension. The section delivered one diagonal and one chord in two short
equations, neither of which required solving a single joint, the economy the
method of sections exists to provide.

\subsection{A wall-mounted bracket crane}

A wall-mounted bracket crane consists of a horizontal jib of length
$L = 2\,\si{\meter}$ pinned to the wall at point $A$ and supported
by a diagonal tie running from a point $B$ on the wall, $1\,\si{
\meter}$ above $A$, to the jib's outer end $C$. A load of $W =
500\,\si{\kilogram} \times 9.81 \,\si{\meter\per\second\squared}
\approx 4{,}900\,\text{N}$ hangs from $C$.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-bracket-crane.tex}

\textbf{FBD of the jib $AC$.} Pin reaction at $A$ ($A_x$, $A_y$);
tension $T$ in the tie along $BC$; load $W$ downward at $C$. The
tie's direction is from $B$ to $C$, which makes an angle below
horizontal of $\arctan(1/2) \approx 26.6^{\circ}$ from $C$; the
tension at $C$ pulls back toward $B$, so its components on the jib
at $C$ are $T \cos\theta$ horizontal (back toward the wall) and
$T \sin\theta$ vertical (upward), where $\theta = 26.6^{\circ}$.

\textbf{Equilibrium.} Taking moments about $A$ eliminates the pin
reactions:
\[
T \sin\theta \cdot L - W \cdot L = 0, \quad
T = \frac{W}{\sin\theta} = \frac{4{,}900}{\sin 26.6^{\circ}}
\approx \frac{4{,}900}{0.447} \approx 11{,}000\,\text{N}.
\]
The tie tension is more than twice the load. Vertical and horizontal
equilibrium then yield the pin reactions:
\[
A_y = W - T \sin\theta = 0, \quad
A_x = T \cos\theta \approx 9{,}800\,\text{N}.
\]
The pin reaction is therefore $9{,}800\,\text{N}$ horizontal, pulling
the jib toward the wall. The pin must be specified for this load,
not for $W$ alone.

The example illustrates two recurring lessons of statics: tie or
cable members can carry forces appreciably larger than the load they
support, and the pin reactions of a crane jib depend on the geometry
of the tie, not only on the load. A crane redesigned to put the tie
closer to horizontal (smaller $\theta$) increases the tie tension
and the pin reaction without changing the load.

\subsection{A beam under combined point and distributed load}

A simply-supported beam spans $L = 8\,\si{\meter}$. It carries a
uniform line load $w_0 = 6\,\text{kN}/\text{m}$ over its full length
plus a point load $P = 40\,\text{kN}$ at $x = 3\,\si{\meter}$ from the
left support. Find the reactions, the maximum bending moment, and its
location. This combination, a deck self-weight plus a concentrated
equipment load, is the everyday case the memorised single-load formulas
do not cover, and it shows the diagram rules carrying their full weight.

\textbf{Reactions.} The total distributed load is $w_0 L = 48\,\text{kN}$
acting at midspan ($x = 4\,\si{\meter}$). Sum moments about the left
support $A$:
\[
R_B \cdot 8 = 48 \cdot 4 + 40 \cdot 3 = 192 + 120 = 312,\quad
R_B = 39\,\text{kN}.
\]
Vertical equilibrium gives $R_A = 48 + 40 - 39 = 49\,\text{kN}$.

\textbf{Shear and the zero crossing.} Just right of $A$ the shear is
$+49\,\text{kN}$. The distributed load sheds shear at $6\,\text{kN}$ per
metre, so over the first $3\,\si{\meter}$ the shear falls to
$49 - 6 \cdot 3 = 31\,\text{kN}$. The point load then drops the shear by
$40\,\text{kN}$ to $-9\,\text{kN}$. The shear therefore crosses zero
inside the point load's jump, at $x = 3\,\si{\meter}$, which by the rule
$dM/dx = V$ locates the maximum moment.

\textbf{Maximum moment.} Evaluate $M$ at $x = 3\,\si{\meter}$ from the
left free body: the left reaction acts at arm $3\,\si{\meter}$, and the
distributed load over the first $3\,\si{\meter}$ is
$6 \cdot 3 = 18\,\text{kN}$ at its own centroid $1.5\,\si{\meter}$ from
the cut:
\[
M_{\max} = 49 \cdot 3 - 18 \cdot 1.5 = 147 - 27 = 120\,\text{kN}\cdot\text{m}.
\]
The routine \texttt{code/shear\_moment.py}, run with this load case,
returns the same peak and confirms the zero crossing at the point load.
The lesson the diagram rules teach is that the maximum moment of a mixed
load sits where the shear crosses zero, which is not midspan and not
under the point load by reflex but at whichever point the running shear
sum reaches zero.

\subsection{A frame with a multi-force member}

The crane resolved into a two-force tie and a pin reaction because the
jib carried its load and its tie pull at the same two points the pin and
the load occupied. A frame whose load lands between a member's supports
forces that member to bend, and the analysis has to disassemble the frame
to see it.

Consider the frame of figure~\ref{fig:vol03ch02:frame-multiforce}. A
horizontal member $AC$ is pinned to a wall at $A = (0, 0)$, runs out to a
free end $C = (3, 0)\,\si{\meter}$ where a load $W = 6\,\text{kN}$ hangs,
and is held up by a tie $BD$ running from a wall anchor $D = (0, 2)$ down
to the midspan point $B = (1.5, 0)$ of $AC$. The tie carries force only
at its two ends, so it is a two-force member and its force lies along its
own axis $DB$. The member $AC$ carries three non-collinear forces (the
pin reaction at $A$, the tie force at $B$, the load at $C$), so it bends
and is a genuine multi-force member.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-frame-multiforce.tex}

\textbf{Tie geometry.} The tie runs from $D = (0, 2)$ to $B = (1.5, 0)$,
a horizontal run of $1.5\,\si{\meter}$ and a vertical drop of
$2\,\si{\meter}$, length $\sqrt{1.5^2 + 2^2} = 2.5\,\si{\meter}$, the
familiar $3$-$4$-$5$ proportions scaled by one half. The unit vector from
$B$ toward $D$ is $(0 - 1.5,\, 2 - 0)/2.5 = (-0.6,\, 0.8)$. A tension in
the tie pulls $B$ toward $D$ along this direction.

\textbf{Tie force by one moment equation.} Isolate $AC$. Take moments
about the pin $A$, which removes both pin-reaction components. The tie
force at $B$ contributes only through its vertical component $0.8\,S$
acting at the $1.5\,\si{\meter}$ arm; the load $W$ acts down at the
$3\,\si{\meter}$ arm:
\[
0.8\,S \cdot 1.5 - W \cdot 3 = 0, \quad
S = \frac{6 \cdot 3}{0.8 \cdot 1.5} = \frac{18}{1.2} = 15\,\text{kN},
\]
positive, so the tie is in tension at $15\,\text{kN}$, two and a half
times the load it holds. The shallow tie angle is the cause, the same
geometry penalty the crane showed.

\textbf{Pin reaction by force balance.} With $S$ known, the two force
equations on $AC$ give the pin reaction. The tie pulls $B$ along
$(-0.6, 0.8)$, so its force on $AC$ is $15 \cdot (-0.6, 0.8) =
(-9,\, +12)\,\text{kN}$. Horizontal equilibrium:
\[
A_x + (-9) = 0, \quad A_x = 9\,\text{kN},
\]
and vertical equilibrium:
\[
A_y + 12 - 6 = 0, \quad A_y = -6\,\text{kN}.
\]
The pin reaction is $(9,\, -6)\,\text{kN}$, magnitude
$\sqrt{9^2 + 6^2} = 10.8\,\text{kN}$. The vertical component is
negative: the pin pulls $AC$ \emph{down}, because the tie's vertical pull
of $12\,\text{kN}$ at $B$ exceeds the $6\,\text{kN}$ load, and the pin
holds the surplus down. An analyst who assumed the pin must push up to
hold a downward load would have the sign backward, and a foundation
detailed only for downward pin load would be wrong for this geometry. The
lesson the frame adds to the crane: disassembly is the price of an
internal force, and the pin reaction of a tied member can reverse
direction from the naive expectation.

\subsection{A beam reaction under a trapezoidal distributed load}

A distributed load that varies along the span is handled by replacing it
with a single resultant at the load's centroid, and the centroid is found
by the composite-area decomposition the chapter has already used for
piecewise-linear distributions. A
simply-supported beam spans $L = 6\,\si{\meter}$ and carries a
trapezoidal line load growing from $w_1 = 4\,\text{kN}/\text{m}$ at the
left support to $w_2 = 10\,\text{kN}/\text{m}$ at the right. Find the
reactions and the maximum bending moment.

\textbf{Resultant by decomposition.} Split the trapezoid into a rectangle
of height $w_1 = 4\,\text{kN}/\text{m}$ over the full span and a triangle
rising from zero at the left to $w_2 - w_1 = 6\,\text{kN}/\text{m}$ at the
right. The rectangle's resultant is
\[
W_r = w_1 L = 4 \cdot 6 = 24\,\text{kN} \text{ at } \bar{x}_r = 3\,\si{\meter},
\]
the midspan. The triangle's resultant is
\[
W_t = \tfrac{1}{2}(w_2 - w_1) L = \tfrac{1}{2}\cdot 6 \cdot 6 = 18\,\text{kN}
\text{ at } \bar{x}_t = \tfrac{2}{3} L = 4\,\si{\meter},
\]
two-thirds of the way toward the deep end, where the triangle's area
piles up. The combined resultant is $W = 24 + 18 = 42\,\text{kN}$, acting
at the area-weighted centroid
\[
\bar{x} = \frac{W_r \bar{x}_r + W_t \bar{x}_t}{W}
= \frac{24 \cdot 3 + 18 \cdot 4}{42}
= \frac{72 + 72}{42} = \frac{144}{42} = 3.43\,\si{\meter}.
\]

\textbf{Reactions.} Sum moments about the left support $A$:
\[
R_B \cdot 6 = W \bar{x} = 42 \cdot 3.43 = 144, \quad R_B = 24\,\text{kN},
\]
and vertical equilibrium gives $R_A = 42 - 24 = 18\,\text{kN}$. The
deeper end carries the larger reaction, as the off-centre centroid
requires.

\textbf{Maximum moment.} The shear runs $V(x) = R_A - \int_0^x w\,ds$,
where $w(x) = w_1 + (w_2 - w_1)x/L = 4 + x$. Integrating,
$V(x) = 18 - 4x - x^2/2$, which crosses zero at
$x = -4 + \sqrt{16 + 36} = \sqrt{52} - 4 \approx 3.21\,\si{\meter}$, just
right of the centroid. The moment there, $M = R_A x - \int_0^x
w(s)(x - s)\,ds$, evaluates to $31.7\,\text{kN}\cdot\text{m}$, the maximum.
The routine \texttt{code/shear\_moment.py}, given a ramp load, returns the
same peak and zero-crossing. The decomposition turned a varying load into
two textbook shapes and an integral into the area-weighting that combines
their centroids.

\subsection{A band brake and the capstan relation}

Friction is the one statics force whose magnitude the equilibrium
equations do not set on their own: it takes whatever value, up to the
limit $\mu N$, that equilibrium requires. The band brake shows the
exponential gain in grip that friction develops when it acts over a
wrapped surface.

A band brake (figure~\ref{fig:vol03ch02:band-brake}) wraps a flexible
band around a drum of radius $r = 0.2\,\si{\meter}$ through a wrap angle
$\beta = 270^{\circ} = 3\pi/2$, with a friction coefficient
$\mu = 0.35$ between band and drum. The two ends of the band carry
different tensions: the side the drum's rotation drags taut carries the
tight tension $T_1$, the other the slack tension $T_2$. Belt friction,
integrated around the wrap, relates them by the capstan (Euler-Eytelwein)
equation
\[
\frac{T_1}{T_2} = e^{\mu\beta} = e^{0.35 \cdot 3\pi/2}
= e^{1.649} = 5.20.
\]
The tight side carries five times the slack side, an exponential payoff
from the wrap angle alone.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-band-brake.tex}

\textbf{Braking torque.} The net tangential drag the band applies to the
drum is the tension difference acting at the drum radius:
\[
M = (T_1 - T_2)\, r.
\]
To deliver a braking torque $M = 300\,\text{N}\cdot\text{m}$, the tension
difference must be $T_1 - T_2 = M/r = 300/0.2 = 1500\,\text{N}$. With
$T_1 = 5.20\, T_2$, the slack tension is
\[
T_2 = \frac{1500}{5.20 - 1} = \frac{1500}{4.20} = 357\,\text{N},
\quad T_1 = 5.20 \cdot 357 = 1857\,\text{N}.
\]

\textbf{Lever force.} The slack end attaches to a brake lever pivoted at
$O$, at a short arm $a = 0.1\,\si{\meter}$ from the pivot, while the
operator presses at a long arm $\ell = 0.6\,\si{\meter}$, the tight end
anchored at the pivot so it exerts no moment. Moment equilibrium of the
lever about $O$:
\[
F \cdot \ell = T_2 \cdot a, \quad
F = \frac{T_2\, a}{\ell} = \frac{357 \cdot 0.1}{0.6} = 59.5\,\text{N}.
\]
A hand force of about $60\,\text{N}$ produces a braking torque of
$300\,\text{N}\cdot\text{m}$. The wrap's exponential and the lever's ratio
multiply: this is why a few turns of rope around a capstan let one sailor
hold a line a winch could barely move, and why a small band brake stops a
heavy drum.

\subsection{A ladder against a wall: the friction limit}

Friction set the band brake's grip; it also sets the moment a leaning
ladder slips. The ladder is the canonical impending-motion problem, where
the friction force sits at its limit $\mu N$ and equilibrium decides whether
the ladder stands. A uniform ladder of length $\ell = 5\,\si{\meter}$ and
weight $W = 200\,\text{N}$ leans against a smooth vertical wall at an angle
$\theta$ to the floor. The floor has friction coefficient $\mu = 0.4$; the
wall is taken frictionless, the conservative idealisation. Find the smallest
angle $\theta$ at which the ladder does not slip, with no one on it, and then
the angle needed to carry an $800\,\text{N}$ person standing at the top.

\textbf{Free body.} Four forces act: the weight $W$ down at the ladder's
midpoint; the wall's normal reaction $N_w$ horizontal at the top (the wall
being frictionless supplies only a normal force); the floor's normal
reaction $N_f$ up at the foot; and the floor's friction $F_f$ horizontal at
the foot, pointing toward the wall to resist the foot sliding out.

\textbf{Equilibrium, ladder alone.} Vertical and horizontal balance give
$N_f = W$ and $F_f = N_w$. Take moments about the foot to remove $N_f$ and
$F_f$. The weight acts at $(\ell/2)\cos\theta$ horizontally from the foot;
the wall reaction acts at $\ell\sin\theta$ vertically:
\[
N_w \,\ell\sin\theta = W\,\frac{\ell}{2}\cos\theta, \quad
N_w = \frac{W}{2}\cot\theta = F_f.
\]
The ladder does not slip while the demanded friction $F_f$ stays under the
available $\mu N_f = \mu W$:
\[
\frac{W}{2}\cot\theta \le \mu W \implies
\tan\theta \ge \frac{1}{2\mu} = \frac{1}{0.8} = 1.25, \quad
\theta \ge 51.3^{\circ}.
\]
A ladder shallower than $51.3^{\circ}$ slips on this floor, the basis of the
four-to-one pitch rule of thumb ($\tan\theta = 4$ gives $76^{\circ}$, an
ample margin).

\textbf{With a person at the top.} A person of weight $Q = 800\,\text{N}$
standing at the very top adds a load at the wall end, the worst position
because it has the longest moment arm about the foot. Moments about the foot
now read
\[
N_w\,\ell\sin\theta = W\,\frac{\ell}{2}\cos\theta + Q\,\ell\cos\theta,
\quad N_w = \Big(\frac{W}{2} + Q\Big)\cot\theta = F_f,
\]
and the vertical balance gives $N_f = W + Q = 1000\,\text{N}$. The no-slip
condition becomes
\[
\Big(\frac{W}{2} + Q\Big)\cot\theta \le \mu (W + Q) \implies
\tan\theta \ge \frac{W/2 + Q}{\mu (W + Q)}
= \frac{100 + 800}{0.4 \cdot 1000} = 2.25, \quad
\theta \ge 66.0^{\circ}.
\]
The climber at the top forces the ladder steeper, from $51^{\circ}$ to
$66^{\circ}$, because the load sits where its moment arm is longest. The
lesson the ladder teaches is that the impending-slip angle depends on
\emph{where} the load is carried, not only on how heavy it is, and the
worst case is the load at the top, the position a painter naturally adopts.

\subsection{A shaft on two bearings: three-dimensional equilibrium}

The planar examples used three equations. A shaft carrying pulleys in
different planes needs the full six, and the analysis shows how the
three-dimensional conditions split into independent planar problems when the
geometry allows. A horizontal shaft (figure~\ref{fig:vol03ch02:shaft-bearings})
turns in two bearings $A$ and $B$ a distance $a + b + c = 0.3 + 0.4 + 0.3 =
1.0\,\si{\meter}$ apart along its axis $z$. Pulley $C$ sits
$a = 0.3\,\si{\meter}$ from bearing $A$ and carries a net belt pull of
$F_C = 3\,\text{kN}$ straight down (in the vertical $xy$ plane, along $-y$).
Pulley $D$ sits a further $b = 0.4\,\si{\meter}$ along and carries a net
belt pull of $F_D = 2\,\text{kN}$ horizontally (in the horizontal plane,
along $+x$). The bearings are short, so each supplies two transverse
reaction components ($A_x, A_y$ and $B_x, B_y$) and no moment.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-shaft-bearings.tex}

\textbf{Why the planes separate.} The two transverse directions $x$ and $y$
do not couple, because no force has components in both and the moment of a
purely-$x$ force about the $x$ axis vanishes, as does the moment of a
purely-$y$ force about the $y$ axis. The six equilibrium equations therefore
split into two independent planar problems, one in the vertical $yz$ plane
driven by $F_C$ and one in the horizontal $xz$ plane driven by $F_D$, plus
the moment balance about the shaft axis that the driving torque satisfies.
Each planar problem is the familiar beam reaction calculation.

\textbf{Vertical plane ($F_C$).} Treat the shaft as a beam in the $yz$
plane carrying $F_C = 3\,\text{kN}$ down at $z = 0.3\,\si{\meter}$ from $A$,
spanning $1.0\,\si{\meter}$ to $B$. Moments about $A$:
\[
B_y \cdot 1.0 = F_C \cdot 0.3 = 0.9, \quad B_y = 0.9\,\text{kN},
\]
and vertical balance gives $A_y = 3 - 0.9 = 2.1\,\text{kN}$, both upward.

\textbf{Horizontal plane ($F_D$).} Treat the shaft as a beam in the $xz$
plane carrying $F_D = 2\,\text{kN}$ along $+x$ at
$z = 0.3 + 0.4 = 0.7\,\si{\meter}$ from $A$. Moments about $A$:
\[
B_x \cdot 1.0 = F_D \cdot 0.7 = 1.4, \quad B_x = 1.4\,\text{kN},
\]
and horizontal balance gives $A_x = 2 - 1.4 = 0.6\,\text{kN}$.

\textbf{Resultant bearing loads.} Each bearing is sized by the magnitude of
its two-component reaction:
\[
|A| = \sqrt{A_x^2 + A_y^2} = \sqrt{0.6^2 + 2.1^2} = 2.18\,\text{kN}, \quad
|B| = \sqrt{B_x^2 + B_y^2} = \sqrt{1.4^2 + 0.9^2} = 1.66\,\text{kN}.
\]
Bearing $A$, nearer the heavier vertical pulley, carries the larger load.
The three-dimensional problem reduced to two planar reaction calculations
the moment the geometry kept the loads in separate planes, and the bearing
selection follows from the resultant magnitudes, not from either component
alone. A designer who sized bearing $A$ for its vertical component
$2.1\,\text{kN}$ and forgot the horizontal $0.6\,\text{kN}$ would
under-specify it by the few percent the resultant adds, small here but
decisive when the two pulleys pull comparably.

\subsection{Case study: a gravity retaining wall carried to a stability verdict}

The cable and shaft balanced applied loads against reactions. A retaining
wall balances applied loads against its own weight, and the analysis is a
contest of moments and forces that decides whether the wall stands. We carry
a gravity retaining wall through the three checks that govern it, the
analysis that sizes a roadway wall or a basement surround.

\textbf{Geometry and loads.} A masonry gravity wall
(figure~\ref{fig:vol03ch02:retaining-wall}) retains soil to a height
$h = 4\,\si{\meter}$. Its cross-section is trapezoidal, $1\,\si{\meter}$ wide
at the top and $B = 3\,\si{\meter}$ at the base, of unit length into the
page, built of masonry of unit weight $\gamma_m = 23\,\text{kN}/\text{m}^3$.
The retained soil has unit weight $\gamma_s = 18\,\text{kN}/\text{m}^3$ and
an active earth-pressure coefficient $K_a = 0.33$. All quantities are taken
per metre of wall length.

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-retaining-wall.tex}

\textbf{The active thrust.} The soil pushes on the back of the wall with a
pressure that grows linearly with depth, summing to a horizontal thrust
\[
P_a = \tfrac12 K_a \gamma_s h^2
= \tfrac12 \cdot 0.33 \cdot 18 \cdot 4^2
= \tfrac12 \cdot 0.33 \cdot 18 \cdot 16 = 47.5\,\text{kN}/\text{m},
\]
acting horizontally at one-third the height, $h/3 = 1.33\,\si{\meter}$ above
the base, the centroid of the triangular pressure diagram.

\textbf{The wall weight and its line of action.} Split the trapezoid into a
rectangle $1\,\si{\meter}$ wide by $4\,\si{\meter}$ tall and a triangle
$2\,\si{\meter}$ wide by $4\,\si{\meter}$ tall. The rectangle's weight is
$W_1 = 23 \cdot 1 \cdot 4 = 92\,\text{kN}$; the triangle's is
$W_2 = 23 \cdot \tfrac12 \cdot 2 \cdot 4 = 92\,\text{kN}$. Measuring the
horizontal distance from the toe (the front bottom edge, at the right of the
figure), the rectangle's centroid sits $0.5\,\si{\meter}$ from its own
front face; placing the rectangle at the back of the wall puts its centroid
$2.5\,\si{\meter}$ from the toe, and the triangle's centroid (one-third of
its width from its vertical back face) lands $1.33\,\si{\meter}$ from the
toe. The total weight is $W = 92 + 92 = 184\,\text{kN}$ at
\[
\bar{x} = \frac{92 \cdot 2.5 + 92 \cdot 1.33}{184}
= \frac{230 + 122}{184} = 1.91\,\si{\meter} \text{ from the toe.}
\]

\textbf{Overturning check.} The wall would topple by rotating about the toe.
The restoring moment is the weight times its arm to the toe; the overturning
moment is the thrust times its height:
\[
M_{\text{resist}} = W\,\bar{x} = 184 \cdot 1.91 = 351\,\text{kN}\cdot\text{m},
\quad
M_{\text{over}} = P_a\,\frac{h}{3} = 47.5 \cdot 1.33 = 63.3\,\text{kN}\cdot\text{m}.
\]
The factor of safety against overturning is the ratio
$351 / 63.3 = 5.5$, comfortably above the customary minimum of about $2$.

\textbf{Sliding check.} The wall would slide forward along its base if the
thrust exceeded the base friction. With a base friction coefficient
$\mu = 0.5$ and the full weight pressing the base down, the available
friction is $\mu W = 0.5 \cdot 184 = 92\,\text{kN}/\text{m}$ against the
thrust $P_a = 47.5\,\text{kN}/\text{m}$, a factor of safety
$92 / 47.5 = 1.9$, near the customary minimum of about $1.5$ and the
governing check of the three.

\textbf{Bearing check.} The base pressure must stay under the soil's
allowable bearing capacity. The resultant of $W$ and $P_a$ strikes the base
at a distance from the toe found by summing moments about the toe:
$\bar{x}_R = (M_{\text{resist}} - M_{\text{over}})/W = (351 - 63.3)/184 =
1.56\,\si{\meter}$, an eccentricity $e = B/2 - \bar{x}_R = 1.5 - 1.56 =
-0.06\,\si{\meter}$ from the base centre, inside the middle third
($|e| < B/6 = 0.5\,\si{\meter}$), so the whole base stays in compression.
The peak base pressure is
\[
p_{\max} = \frac{W}{B}\Big(1 + \frac{6|e|}{B}\Big)
= \frac{184}{3}\Big(1 + \frac{6 \cdot 0.06}{3}\Big)
= 61.3 \cdot 1.12 = 68.7\,\text{kPa},
\]
which a competent founding soil (allowable bearing near
$200\,\text{kPa}$) carries with ample margin.

\textbf{The verdict.} The wall is safe against all three modes, with sliding
the governing check at a factor of safety of $1.9$. The reading a designer
takes away is that a gravity wall's stability is bought with weight and base
width: widening the base raises the restoring moment and the base friction
together while lowering the peak bearing pressure, so the base width is the
single dimension that trades against the retained height across all three
checks at once.

\begin{estimation}
\textbf{Estimate the bolt shear capacity needed to hang the bracket
crane's load through a single bolt at pin $A$.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The horizontal pin reaction came out near $9{,}800\,\text{N}$, so a
single bolt at $A$ must carry roughly $10\,\text{kN}$ in shear. Order
the bolt by the rule $V_u \approx 0.6 F_u A_b$. For a common
grade with $F_u \approx 800\,\text{MPa} = 8 \times 10^{8}\,\text{Pa}$,
the area needed for ultimate capacity equal to the load is
\[
A_b \approx \frac{V_u}{0.6 F_u} =
\frac{10^{4}}{0.6 \times 8 \times 10^{8}}
\approx 2.1 \times 10^{-5}\,\si{\meter\squared}
= 21\,\si{\milli\meter\squared}.
\]
A bolt of that area has diameter $d = \sqrt{4 A_b / \pi}
\approx 5\,\si{\milli\meter}$. That is the bolt at the breaking point;
applying a factor of safety of three on a static connection pushes the
required area to $63\,\si{\milli\meter\squared}$ and the diameter to
about $9\,\si{\milli\meter}$, so a standard $\text{M10}$ bolt is the
estimate. The postmortem: the load is $4{,}900\,\text{N}$ but the bolt
is sized for $9{,}800\,\text{N}$, twice the load, because the geometry
of the tie doubled the pin reaction. Sizing the bolt from the hung
weight rather than the computed reaction would have under-specified it
by a factor of two, the recurring trap this chapter names.
\end{estimation}

\section{Failure: pin connections, bolted joints, missed load paths}
\pathtag{core}

Pin and bolt connections are the points at which statics's
clean line diagrams meet the structural reality of real fasteners.
A truss member that the analysis treats as a perfect two-force
element is, at each end, a hole in a steel plate through which a
bolt passes. The bolt transfers the member's axial force to the
adjoining plate by shear in the bolt and bearing between the bolt
shank and the hole's edge. The plate transfers the force into the
joint by tension in the net section between the hole and the
plate's edge. The joint's failure modes are correspondingly several,
drawn together in figure~\ref{fig:vol03ch02:bolt-failure-modes}:

\begin{itemize}[leftmargin=*]
  \item \engterm{Bolt shear.} The bolt fails in shear across its
    cross-section. Capacity scales with bolt cross-sectional area
    and the shear strength of the bolt material. A bolt of nominal
    diameter $d$ in single shear has area $\pi d^2 / 4$; the shear
    capacity per bolt is approximately $0.6 F_u (\pi d^2 / 4)$ for
    a high-strength bolt with ultimate tensile strength $F_u$.
  \item \engterm{Bearing.} The bolt's shank bears on the hole's
    edge in the plate. If the plate is thin or low-strength, the
    edge of the hole deforms; the bolt elongates the hole until
    the plate tears. Capacity scales with bolt diameter, plate
    thickness, and plate ultimate tensile strength.
  \item \engterm{Net-section tension.} The plate, weakened by the
    hole, fails in tension across the net section (the gross
    section minus the hole). Capacity is the net area times the
    plate's ultimate tensile strength. The reduction from gross to
    net section is the engineering reason a hole is more critical
    than its diameter alone suggests.
  \item \engterm{Block shear.} A combination of shear failure
    along two planes parallel to the load and tension failure
    perpendicular, in the plate's region behind the bolt. Block
    shear is the dominant failure mode for many bolted-end
    connections in tension members; it is the failure mode that the
    NBS investigation of the Hyatt Regency walkway collapse named
    as the mechanism by which the modified upper connection tore
    out of the box-beam's lower flange \cite{acc:nbs-hyatt-1982}.
  \item \engterm{Bolt tension and preload loss.} High-strength
    bolted joints rely on bolt preload (clamping force) to develop
    friction at the joint interface and to carry transverse loads
    by friction rather than by bolt shear. Preload loss from
    vibration, gasket creep, thermal cycling, or under-torquing
    converts the joint from friction-grip to bearing-on-bolt, with
    a corresponding drop in capacity and increase in fatigue
    sensitivity. Shigley's design handbook covers the preload
    analysis in detail \cite[ch.~8]{text:shigley2020}.
\end{itemize}

\input{volumes/03-force/ch02-statics-of-rigid-bodies/figures/fig-bolt-failure-modes.tex}

The chapter's discipline is statics, and statics catches each of
these failure modes only if the analyst draws the right free-body
diagrams. A connection analysed as a pin (two reaction components,
no moment) hides the bolt's shear and bearing checks unless the
engineer steps inside the pin and asks how the force gets across.
A connection analysed as a stack of bolts in pure friction hides
the bolt-shear check unless the engineer asks what happens when the
friction slips.

Chapter 1's failure section introduced the Hyatt Regency walkway
collapse as the canonical case of a free-body diagram that hid the
real load path. The chapter-2 lesson is the same failure read from
the joint side: the modified upper box-beam connection's tear-out
mechanism was a joint failure that no joint-level FBD had been
drawn to check. The structural engineer of record's signed shop
drawing showed the modification's geometry but not the new
connection's free body. The bolt-shear, bearing, and block-shear
checks for the doubled load were never performed
\cite{acc:nbs-hyatt-1982}. The mechanism of the failure (block shear
through the box-beam's lower flange) is the canonical bolted-joint
failure mode this section names.

The general principle that closes the chapter:

\begin{principle}[Joints are not pins]
A connection idealised as a pin in the global FBD is, in reality, a
fastener-and-plate assembly with its own failure modes (bolt shear,
bearing, net-section tension, block shear, preload loss). Every
joint in a statically analysed structure requires its own
free-body-diagram check against each of those modes at the
connection level. The global statics does not catch a joint
failure; the joint FBD does.
\end{principle}

We return to bolted-joint detail design in chapter 8 of this volume
(in the context of beams, columns, and frames), in chapter 8 of
Volume V (in the context of material-failure modes), and in chapter
8 of Volume VIII (in the context of machine elements).

\section{Project}\pathtag{core}

\begin{project}[Design, build, and load-test a wooden truss]
The reader designs, builds, and tests a small wooden Howe or Pratt
truss to failure under controlled bench load. The deliverable is a
forty-page engineering report with FBDs, calculations, the built
truss (or photographs thereof), the test data, and a post-test
analysis.

\textbf{Track}. Physical build plus analysis. \textbf{Hazard
class}: low. Eye protection during failure testing is required.
Loads at failure are typically under $200\,\text{N}$; no risk of
flying debris with proper containment of the test rig.

\textbf{Design}. A simply-supported truss of span $400\,\si{
\milli\meter}$ to $800\,\si{\milli\meter}$, depth one-fifth to
one-tenth of the span, made from straight wooden members of
$5\,\si{\milli\meter} \times 5\,\si{\milli\meter}$ to $10\,\si{
\milli\meter} \times 10\,\si{\milli\meter}$ cross-section, joined at
the bottom chord joints by short pieces of pin or by glued plywood
gusset plates. Six to ten bays. The design target is a maximum
midspan load of at least $100\,\si{\newton}$ before failure.

\textbf{Analysis}. Before construction, the reader produces:

\begin{itemize}[leftmargin=*]
  \item The truss geometry, dimensioned.
  \item FBD of the whole truss, with reactions calculated for a
    centred midspan point load.
  \item Joint-by-joint or method-of-sections analysis giving the
    force in each member.
  \item Identification of the most-loaded member (compression and
    tension separately) and a prediction of which failure mode
    (buckling, crushing, tension fracture, joint slip) is expected
    to govern.
  \item A predicted failure load with stated uncertainty (factor of
    $1.5$ to $3$ is acceptable for a first-time builder).
\end{itemize}

\textbf{Construction}. The reader builds the truss from the
dimensioned design, photographing each joint as constructed.
Departures from the design (member length errors, joint
misalignment) are recorded.

\textbf{Test}. The truss is simply supported on two blocks. Load is
applied at midspan by a hanging weight (water jug, bag of sand,
known masses) or by a screw-driven plunger with a strain-gauge or
spring scale. Load is incremented in $10\,\si{\newton}$ steps;
deflection is measured at each step; the reader watches for joint
slip, member buckling, or audible cracking. The load at first
failure is recorded; the failure mode is photographed.

\textbf{Analysis after test}. The reader writes a $1{,}500$ to
$3{,}000$-word post-test report addressing:

\begin{itemize}[leftmargin=*]
  \item Predicted failure load versus measured failure load;
    discrepancy in factors.
  \item Predicted failure mode versus actual failure mode; if
    different, what was missed in the prediction.
  \item Which joint or member governed; what would have to change
    to move the failure to a different location.
  \item One sentence per truss member naming the force the analysis
    predicted and (where measurable) the qualitative behaviour
    observed.
\end{itemize}

\textbf{Deliverable}. A bound or PDF report with the analysis, the
design drawings, photographs of the truss before, during, and after
testing, the test data, and the post-test analysis.

\textbf{Time}. Four to eight hours of design and analysis, two to
four hours of construction, one to two hours of testing, four to
eight hours of post-test analysis and write-up.
\end{project}

\section{Exercises}

The exercises mix calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment. Solutions are
provided for calculation and derivation exercises; estimation
exercises receive published reference estimates with the editor's
working; design, diagnosis, failure-analysis, and judgment exercises
receive rubrics rather than single answers. Exercises are tagged
with their reader-path tier.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A simply-supported beam of span $6\,\si{\meter}$ carries a uniform
line load of $5\,\text{kN}/\text{m}$. Find the support reactions, the
maximum bending moment, and the location of the maximum.

\paragraph{Solution sketch.} Total load $W = w_0 L = 5 \times 6 =
30\,\text{kN}$, symmetric, so each reaction is $R = W/2 =
15\,\text{kN}$. The maximum moment of a uniformly loaded
simply-supported beam is $M_{\max} = w_0 L^2 / 8 = 5 \times 36 / 8 =
22.5\,\text{kN}\cdot\text{m}$ at midspan ($x = 3\,\si{\meter}$), where
the shear $V(x) = R - w_0 x = 15 - 5x$ crosses zero. The routine
\texttt{code/shear\_moment.py} returns the same peak.
\end{exercise}

\begin{exercise}
A simply-supported beam of span $4\,\si{\meter}$ carries a centred
point load of $10\,\text{kN}$ plus a uniform line load of $2\,
\text{kN}/\text{m}$. Sketch the shear and bending moment diagrams,
and find the maximum bending moment.

\paragraph{Solution sketch.} Superpose the two cases. The point load
contributes reactions of $5\,\text{kN}$ each and a midspan moment of
$PL/4 = 10 \times 4 / 4 = 10\,\text{kN}\cdot\text{m}$. The uniform load
contributes reactions of $w_0 L / 2 = 2 \times 4 / 2 = 4\,\text{kN}$
each and a midspan moment of $w_0 L^2 / 8 = 2 \times 16 / 8 =
4\,\text{kN}\cdot\text{m}$. Both maxima sit at midspan, so they add:
$R = 9\,\text{kN}$ each and $M_{\max} = 10 + 4 =
14\,\text{kN}\cdot\text{m}$ at $x = 2\,\si{\meter}$. The shear diagram
slopes from $+9$ down through $+5$ just left of midspan (losing
$2\,\text{kN}/\text{m}$), jumps by $-10$ at the load to $-5$, and slopes
to $-9$ at the right support; it crosses zero at midspan, confirming the
moment peak there. The moment diagram is a parabola with a kink at
midspan.
\end{exercise}

\begin{exercise}
A Pratt truss of span $L$, height $h$, and $n$ equal bays carries
equal joint loads $P$ at each interior bottom joint. State the
midspan top-chord force as a function of $L$, $h$, $n$, and $P$.
Verify by checking the case $L = 12\,\si{\meter}$, $h = 2\,\si{\meter}$,
$n = 6$, $P = 20\,\text{kN}$ against the worked example.

\paragraph{Solution sketch.} Treat the truss as a simply-supported
beam carrying $n-1$ equal interior loads $P$ at the bay stations, with
bay width $a = L/n$. By symmetry each reaction is $(n-1)P/2$. Summing
moments of the left half about midspan for an even number of bays gives
the midspan moment $M_{\max} = n^2 P a / 8 = n P L / 8$. The chord force
resists this moment as a couple over the depth $h$:
\[
F_{\text{chord}} = \frac{M_{\max}}{h} = \frac{n P L}{8 h}.
\]
Check: $n = 6$, $L = 12$, $h = 2$, $P = 20$ gives
$F = (6)(20)(12)/(8 \times 2) = 1440/16 = 90\,\text{kN}$, matching the
worked example exactly. The top chord carries this in compression and
the bottom chord in tension.
\end{exercise}

\begin{exercise}
A cantilever beam of length $L$ has a fixed support at one end and
carries a uniform line load $w_0$ over its full length. Find the
fixed-end reactions (vertical force and moment) and the maximum
bending moment.

\paragraph{Solution sketch.} The total load $W = w_0 L$ acts at the
centroid $L/2$ from the fixed end. Vertical equilibrium gives the
fixed-end reaction $R = w_0 L$ upward. Moment equilibrium about the
fixed end gives the reaction moment $M_{\text{fix}} = w_0 L \cdot (L/2)
= w_0 L^2 / 2$. The bending moment is maximum at the fixed end and
equals the reaction moment in magnitude, $M_{\max} = w_0 L^2 / 2$
(hogging, the beam concave down). This is four times the
simply-supported midspan moment $w_0 L^2/8$ for the same span and load,
the quantitative reason cantilevers are the demanding case.
\end{exercise}

\begin{exercise}
A rectangular plate of width $b = 100\,\si{\milli\meter}$ and depth
$h = 50\,\si{\milli\meter}$ has a circular hole of diameter $d =
20\,\si{\milli\meter}$ at its centroid. Find the centroid of the
plate-with-hole and its second moment of area about the centroidal
axis perpendicular to $h$.

\paragraph{Solution sketch.} The hole sits at the centroid of the
gross rectangle, so by symmetry the centroid of the plate-with-hole is
unchanged, at the geometric centre. For the second moment about the
horizontal centroidal axis (bending about the axis parallel to $b$),
subtract the hole's contribution from the gross:
\[
I = \frac{b h^3}{12} - \frac{\pi d^4}{64}
= \frac{0.1 \times 0.05^3}{12} - \frac{\pi (0.02)^4}{64}.
\]
The gross term is $0.1 \times 1.25 \times 10^{-4} / 12 =
1.042 \times 10^{-6}\,\si{\meter\tothe{4}}$. The hole term is
$\pi \times 1.6 \times 10^{-7} / 64 = 7.85 \times 10^{-9}\,
\si{\meter\tothe{4}}$, under one percent of the gross. So
$I \approx 1.034 \times 10^{-6}\,\si{\meter\tothe{4}}$. The hole
removes little stiffness because it sits on the neutral axis, where the
material was contributing least; the same hole near the extreme fibre
would cost far more. The routine
\texttt{code/section\_properties.py} reproduces the result with a
negative-area rectangle approximating the hole.
\end{exercise}

\begin{exercise}
A standard rolled I-section has flanges $200\,\si{\milli\meter} \times
20\,\si{\milli\meter}$ and a web $200\,\si{\milli\meter} \times 10\,
\si{\milli\meter}$ (total depth $240\,\si{\milli\meter}$). Compute
the section's second moment of area about its centroidal axis
parallel to the flanges. Compare to a solid rectangular section of
the same total area.

\paragraph{Solution sketch.} By symmetry the centroid is at
mid-height. Decompose into web plus two flanges and apply the
parallel-axis theorem. Web: $I_w = b_w h_w^3 / 12 =
0.01 \times 0.2^3 / 12 = 6.67 \times 10^{-6}\,\si{\meter\tothe{4}}$,
with no transfer term (web centred). Each flange:
$I_{f,\text{self}} = 0.2 \times 0.02^3/12 = 1.33 \times 10^{-7}$, plus a
transfer term $A_f d^2$ where $A_f = 0.2 \times 0.02 = 4 \times
10^{-3}\,\si{\meter\squared}$ and $d = (0.2 + 0.02)/2 = 0.11\,\si{
\meter}$, giving $A_f d^2 = 4 \times 10^{-3} \times 0.0121 =
4.84 \times 10^{-5}$. Two flanges contribute
$2(1.33 \times 10^{-7} + 4.84 \times 10^{-5}) = 9.71 \times 10^{-5}$.
Total $I = 6.67 \times 10^{-6} + 9.71 \times 10^{-5} \approx
1.04 \times 10^{-4}\,\si{\meter\tothe{4}}$, matching
\texttt{code/section\_properties.py}. The total area is
$2(4 \times 10^{-3}) + 0.01 \times 0.2 = 1.0 \times 10^{-2}\,\si{
\meter\squared}$; a solid square of that area is $0.1\,\si{\meter}$ on a
side with $I = 0.1^4/12 = 8.3 \times 10^{-6}\,\si{\meter\tothe{4}}$. The
I-section is about twelve times stiffer for the same material, the whole
case for the shape.
\end{exercise}

\begin{exercise}
A wall-mounted bracket has a horizontal jib of length $1.5\,\si{
\meter}$ and a tie at $30^{\circ}$ above horizontal connecting the
wall to the jib's outer end. The jib carries a $200\,\si{\kilogram}$
load at its outer end. Find the tie tension and the pin reaction at
the wall.

\paragraph{Solution sketch.} The load is $W = 200 \times 9.81 =
1{,}962\,\text{N}$. The tie makes $\theta = 30^{\circ}$ with horizontal
and attaches at the jib's outer end. Take moments about the wall pin
$A$: the load acts at arm $L = 1.5\,\si{\meter}$ and the tie's vertical
component $T \sin\theta$ acts at the same arm, so
$T \sin 30^{\circ} \cdot L = W \cdot L$, giving
$T = W / \sin 30^{\circ} = 1{,}962 / 0.5 = 3{,}924\,\text{N}$. The tie
tension is twice the load. Vertical equilibrium gives the pin's vertical
reaction $A_y = W - T \sin\theta = 0$; horizontal equilibrium gives
$A_x = T \cos 30^{\circ} = 3{,}924 \times 0.866 = 3{,}398\,\text{N}$,
pulling the jib toward the wall. The pin reaction is therefore about
$3.4\,\text{kN}$, horizontal, larger than the hung weight.
\end{exercise}

\begin{exercise}
A two-force member runs from $A = (0, 0)$ to $B = (3, 4)$ (in
metres) and carries a tensile force of $50\,\text{kN}$. State the
$x$- and $y$-components of the force the member exerts on the joint
at $B$.

\paragraph{Solution sketch.} The member length is
$\sqrt{3^2 + 4^2} = 5\,\si{\meter}$, so the unit vector from $A$ to
$B$ is $(3/5, 4/5) = (0.6, 0.8)$. A tension member pulls the joint at
$B$ back toward $A$, that is, along $-\hat{u} = (-0.6, -0.8)$. The force
on joint $B$ is therefore $50 \times (-0.6, -0.8) = (-30, -40)\,\text{kN}$:
$F_x = -30\,\text{kN}$, $F_y = -40\,\text{kN}$. The signs encode that a
member in tension pulls its end joints inward toward the member's
interior.
\end{exercise}

\begin{exercise}
A vertical wall of height $h = 5\,\si{\meter}$ retains water at
maximum height. Find the horizontal force per unit length of wall
exerted by the water on the wall, and the height above the base at
which the equivalent point force acts. Water density $\rho =
1000\,\si{\kilogram\per\meter\cubed}$, $g = 9.81\,\si{\meter\per\second
\squared}$.

\paragraph{Solution sketch.} Hydrostatic pressure grows linearly with
depth, $p(z) = \rho g z$, reaching $p_{\max} = \rho g h$ at the base.
Per unit length of wall this is a triangular line load. The resultant
per unit length is the area of the pressure triangle:
\[
F = \tfrac12 \rho g h^2 = \tfrac12 \times 1000 \times 9.81 \times 25
= 1.23 \times 10^{5}\,\text{N/m} = 123\,\text{kN/m}.
\]
The centroid of a triangle sits one-third of its height from the base,
so the resultant acts at $h/3 = 5/3 \approx 1.67\,\si{\meter}$ above the
base. The one-third rule is the recurring fact of hydrostatic loading:
the resultant of a triangular pressure distribution always acts at the
third-point nearest the deepest water.
\end{exercise}

\begin{exercise}
A bolt of nominal diameter $16\,\si{\milli\meter}$ joins two plates
in single shear. The bolt's ultimate tensile strength is $F_u =
830\,\text{MPa}$. Estimate the bolt's shear capacity using the
approximation $V_u = 0.6 F_u A_b$ with $A_b$ the gross
cross-sectional area. State the assumption that makes this an
estimate rather than a design value.

\paragraph{Solution sketch.} Gross area $A_b = \pi d^2 / 4 =
\pi (0.016)^2 / 4 = 2.01 \times 10^{-4}\,\si{\meter\squared} =
201\,\si{\milli\meter\squared}$. Shear capacity
$V_u = 0.6 F_u A_b = 0.6 \times 830 \times 10^{6} \times
2.01 \times 10^{-4} = 1.00 \times 10^{5}\,\text{N} \approx
100\,\text{kN}$. The result is an estimate, not a design value, for
several reasons: it uses the gross area rather than the reduced threaded
(tensile-stress) area that governs if the shear plane crosses the
threads; it omits the resistance factor (or factor of safety) that code
applies to reach an allowable load; and the $0.6$ coefficient is an
approximation to the shear-to-tension strength ratio rather than a
measured property of this lot. A design value would use the threaded
area, the code resistance factor, and the specified rather than nominal
strength.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Prove that the moment of a force about any point in a plane depends
only on the force's line of action, not on the point of application
along that line. State the analogous result in three dimensions.

\paragraph{Solution sketch.} Let the force $\vec{F}$ act at a point
$P$, and let $O$ be the moment centre. The moment is $\vec{M}_O =
\vec{r}_{OP} \times \vec{F}$. Slide the application point to $P'$ along
the line of action, so $\vec{r}_{OP'} = \vec{r}_{OP} + s\vec{F}/|\vec{F}|$
for some scalar $s$ (the displacement is parallel to $\vec{F}$). Then
$\vec{M}_O' = \vec{r}_{OP'} \times \vec{F} = \vec{r}_{OP} \times \vec{F}
+ (s/|\vec{F}|)(\vec{F} \times \vec{F})$. The cross product of a vector
with itself vanishes, so $\vec{M}_O' = \vec{M}_O$. The moment is
unchanged: this is the principle of transmissibility, and it is why a
force may be represented by its magnitude, direction, and line of
action alone. In three dimensions the same argument holds verbatim with
$\vec{M}_O = \vec{r} \times \vec{F}$ a vector; the moment is invariant
to sliding the force along its line of action, and the moment vector is
perpendicular to the plane containing $O$ and the line.
\end{exercise}

\begin{exercise}
Derive the centroid of a right triangle with legs $a$ and $b$ along
the coordinate axes. Show that the centroid is at $(a/3, b/3)$.

\paragraph{Solution sketch.} Place the right angle at the origin, with
legs along the axes and the hypotenuse running from $(a, 0)$ to
$(0, b)$. The bounding line is $y = b(1 - x/a)$. The area is $A = ab/2$.
The first moment about the $y$-axis:
\[
\int_A x\, dA = \int_0^a x \, y(x)\, dx
= \int_0^a x\, b\Big(1 - \frac{x}{a}\Big) dx
= b\Big[\frac{a^2}{2} - \frac{a^2}{3}\Big] = \frac{a^2 b}{6}.
\]
Then $\bar{x} = (a^2 b / 6)/(ab/2) = a/3$. By the symmetric argument
with the roles of $x$ and $y$ exchanged, $\bar{y} = b/3$. The centroid
of a triangle sits one-third of the way from each side toward the
opposite vertex, the geometric fact behind the two-thirds-span rule for
triangular distributed loads.
\end{exercise}

\begin{exercise}
Derive the second moment of area of a rectangular section of width
$b$ and depth $h$ about its centroidal axis parallel to $b$. Show
that $I = bh^3 / 12$. State explicitly which axis the formula
refers to.

\paragraph{Solution sketch.} Take the centroidal axis parallel to the
width $b$ (the axis about which the section bends when the depth $h$ is
vertical). Measure $y$ from this axis, so $y$ runs from $-h/2$ to
$+h/2$, and a strip of area $dA = b\, dy$. Then
\[
I = \int_A y^2\, dA = \int_{-h/2}^{h/2} y^2\, b\, dy
= b\Big[\frac{y^3}{3}\Big]_{-h/2}^{h/2}
= b\cdot \frac{2}{3}\Big(\frac{h}{2}\Big)^3 = \frac{b h^3}{12}.
\]
The cube on $h$ is the whole point: the formula refers to bending about
the axis perpendicular to the depth direction, and stiffness scales with
the cube of the dimension in the plane of bending. About the orthogonal
centroidal axis the roles swap and $I = h b^3 / 12$.
\end{exercise}

\begin{exercise}
Prove the parallel-axis theorem: for a planar region with second
moment of area $I_C$ about a centroidal axis, the second moment
about a parallel axis at perpendicular distance $d$ is $I = I_C + A
d^2$.

\paragraph{Solution sketch.} Let the centroidal axis be $y = 0$ and the
parallel axis be $y = -d$, so distance to the parallel axis is $y + d$.
The second moment about the parallel axis is
\[
I = \int_A (y + d)^2\, dA
= \int_A y^2\, dA + 2d \int_A y\, dA + d^2 \int_A dA.
\]
The first integral is $I_C$ by definition. The middle integral is the
first moment of area about the centroidal axis, which vanishes by the
definition of the centroid. The last integral is $A$. Therefore
$I = I_C + A d^2$. The cross term dropping out is the only subtlety, and
it drops out precisely because the reference axis passes through the
centroid; the theorem fails if the starting axis is not centroidal.
\end{exercise}

\begin{exercise}
Derive the differential relations $dV/dx = -w(x)$ and $dM/dx = V(x)$
for a planar beam under distributed load $w(x)$. State the sign
conventions explicitly.

\paragraph{Solution sketch.} Take $x$ along the beam, $w(x)$ the
distributed load per unit length positive downward, $V$ positive when it
shears the right face downward, and $M$ positive sagging. Isolate an
element of length $dx$ between $x$ and $x + dx$, carrying $V$ and $M$ on
the left face and $V + dV$, $M + dM$ on the right. Vertical equilibrium:
$V - (V + dV) - w\, dx = 0$, which gives $dV = -w\, dx$, hence
$dV/dx = -w(x)$. Moment equilibrium about the right face:
$M + V\, dx - (M + dM) - w\, dx \cdot (dx/2) = 0$. The last term is
second order in $dx$ and drops in the limit, leaving $V\, dx = dM$, hence
$dM/dx = V(x)$. Differentiating the second and substituting the first
gives $d^2M/dx^2 = -w(x)$. The sign conventions are the load of the
result: a different shear convention flips the sign on the first
relation, which is why every beam analysis must state its conventions
before drawing a single diagram.
\end{exercise}

\begin{exercise}[\textit{deferred to Chapter 8}]
A simply-supported beam of span $L$ carries a uniform line load
$w_0$. Derive the deflection at midspan using the moment-area
theorems. Identify the second moment of area of the cross-section
and the elastic modulus as the only material and geometric
quantities needed. Chapter 8 of this volume develops the
moment-area machinery.

\paragraph{Solution sketch.} The deflection is deferred because it
requires the elastic relation $M = EI\, \kappa$ that chapter 8 develops;
statics alone gives the moment diagram but not the displacement. The
moment-area route, for completeness: the $M/EI$ diagram of a uniformly
loaded simply-supported beam is a parabola peaking at $w_0 L^2/(8EI)$.
The second moment-area theorem gives the midspan deflection as the
moment of the $M/EI$ area between support and midspan, taken about the
midspan, worked out to the standard result
\[
\delta_{\text{mid}} = \frac{5 w_0 L^4}{384\, EI}.
\]
The only material and geometric inputs are the elastic modulus $E$ and
the second moment of area $I$; the statics fixes everything else. The
$L^4$ dependence is the reason deflection, not strength, governs the
design of long slender spans.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the maximum bending moment in a $4\,\si{\meter}$
simply-supported wooden floor joist carrying a uniform live load
representative of a residential bedroom. State assumptions and
factor-of-three bracket.

\paragraph{Solution sketch.} Residential live load runs near
$1.5\text{ to }2\,\text{kPa}$. Assume joists at $400\,\si{\milli\meter}$
spacing, so each joist carries a tributary strip $0.4\,\si{\meter}$
wide: $w_0 \approx 2 \times 10^{3} \times 0.4 = 800\,\text{N/m}$. The
maximum moment of a simply-supported joist is $w_0 L^2/8 =
800 \times 16 / 8 = 1{,}600\,\text{N}\cdot\text{m} \approx
1.6\,\text{kN}\cdot\text{m}$. Bracketing the live load and spacing over
a factor of three each way puts the estimate between roughly
$0.6$ and $4\,\text{kN}\cdot\text{m}$. The reference value sits near
$1.5\,\text{kN}\cdot\text{m}$ for a code bedroom load, inside the
bracket. A dead-load allowance for the floor itself would add perhaps
twenty percent.
\end{exercise}

\begin{exercise}
Estimate the tension in the main rope of a swing supporting an adult
swinging at the bottom of an arc, given the rope length and the
swing speed at the bottom. State which force component is the
weight and which is the centripetal contribution.

\paragraph{Solution sketch.} At the bottom of the arc the rope tension
supplies both the weight and the centripetal force: $T = mg + m v^2 / r$.
Take an adult of $m = 75\,\si{\kilogram}$, rope length $r = 3\,\si{
\meter}$, and a brisk bottom-of-arc speed $v \approx 4\,\si{\meter\per
\second}$. The weight term is $mg = 75 \times 9.81 \approx 740\,\text{N}$;
the centripetal term is $m v^2 / r = 75 \times 16 / 3 \approx 400\,
\text{N}$. The tension is about $1{,}100\,\text{N}$, roughly $1.5$ times
the rider's weight. The weight term is the static part, present even at
rest; the centripetal term is the dynamic surcharge from the curved
path. The swing rope, like the crane tie, carries more than the load it
holds, here because the motion adds to gravity rather than because of
geometry.
\end{exercise}

\begin{exercise}
Estimate the force on the keel of a small sailboat held at heel
by wind on a tight beat. State assumptions about boat geometry,
wind speed, and sail area; bracket the answer.

\paragraph{Solution sketch.} The keel's side force balances the sail's
side force in steady heel, so estimate the aerodynamic side force on the
sail. Dynamic pressure at wind speed $U$ is $q = \tfrac12 \rho_{\text{air}}
U^2$; with $\rho_{\text{air}} = 1.2\,\si{\kilogram\per\meter\cubed}$ and
$U \approx 8\,\si{\meter\per\second}$ (a stiff breeze), $q \approx 38\,
\text{Pa}$. For a sail area $A \approx 15\,\si{\meter\squared}$ and an
effective side-force coefficient near $1$, the side force is $q A
\approx 570\,\text{N}$. The keel must supply roughly this side force in
the water, so the lateral hydrodynamic load on the keel is of order
$0.5\text{ to }1\,\text{kN}$. Bracketing wind speed and sail area over a
factor of three each spreads the estimate from about $0.2$ to
$3\,\text{kN}$. The heeling moment, keel-to-sail centre-of-effort arm
times this force, is the design driver for the hull and rigging, not the
force alone.
\end{exercise}

\begin{exercise}
Estimate the bolt diameter required to secure a $5\,\si{\meter}$
extension ladder's top against the wind acting on its sail area.
State the ladder geometry, the wind speed, and the assumed bolt
strength.

\paragraph{Solution sketch.} Take the ladder's projected sail area as
its two rails plus rungs, roughly $0.3\,\si{\meter}$ wide over
$5\,\si{\meter}$, so $A \approx 1.5\,\si{\meter\squared}$. A storm wind
$U \approx 25\,\si{\meter\per\second}$ gives dynamic pressure
$q = \tfrac12 \times 1.2 \times 625 \approx 375\,\text{Pa}$ and a drag
force $F \approx C_d q A \approx 1.2 \times 375 \times 1.5 \approx
675\,\text{N}$, call it $700\,\text{N}$. A single retaining bolt at the
top must carry this in shear. With $V_u = 0.6 F_u A_b$ and
$F_u \approx 400\,\text{MPa}$ for a plain steel bolt, the area for
ultimate capacity equal to the load is
$A_b = 700/(0.6 \times 4 \times 10^{8}) \approx 3\,\si{\milli\meter
\squared}$, a $2\,\si{\milli\meter}$ bolt. Applying a factor of safety
of five against gusts and stress concentration lifts the required area
to $15\,\si{\milli\meter\squared}$ and the diameter to about
$4.5\,\si{\milli\meter}$, so an $\text{M5}$ or $\text{M6}$ bolt. The
bracket spans wind speed and area over a factor of three each, which
swings the force by an order of magnitude and the diameter by about
three.
\end{exercise}

\begin{exercise}
Estimate the bending moment at the base of a $2\,\si{\meter}$
free-standing road sign in a $30\,\si{\meter\per\second}$ wind.
State assumptions about the sign's frontal area and drag
coefficient.

\paragraph{Solution sketch.} Take a square sign $2\,\si{\meter}$ on a
side, so frontal area $A = 4\,\si{\meter\squared}$, mounted with its
centre at height $H \approx 3\,\si{\meter}$. Dynamic pressure
$q = \tfrac12 \times 1.2 \times 30^2 = 540\,\text{Pa}$. A flat plate
normal to the flow has drag coefficient $C_d \approx 1.2$, so the wind
force is $F = C_d q A = 1.2 \times 540 \times 4 \approx 2{,}600\,
\text{N}$. The bending moment at the base is the force times the height
of its line of action (the sign's centroid):
$M_{\text{base}} = F \times H \approx 2{,}600 \times 3 \approx
7{,}800\,\text{N}\cdot\text{m} \approx 8\,\text{kN}\cdot\text{m}$. The
post and its foundation must resist this overturning moment, which is
why sign posts are sized by wind moment, not by the sign's weight. A
factor-of-three bracket on area and wind puts the moment between roughly
$3$ and $25\,\text{kN}\cdot\text{m}$.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a numerical truss solver for an arbitrary planar truss
specified by joint coordinates, member connectivity, applied loads,
and support types. Output: member forces and support reactions.
Verify on the worked-example Pratt truss; verify on a textbook
Warren truss.

\paragraph{Solution sketch.} The reference implementation is
\texttt{code/method\_of\_joints.py}: assemble the $2j$ joint-equilibrium
equations into a matrix whose columns are the $m$ member unit vectors and
the $r$ reaction components, then solve the linear system $A\vec{x} =
\vec{b}$ where $\vec{b}$ holds the negated applied loads. A tension
member contributes its unit vector to one joint and its negative to the
other. The solver checks determinacy ($m + r = 2j$) before solving and
raises if the system is singular, which flags a mechanism or a
mislabelled support. Verification on the Pratt truss reproduces the
worked example: reactions $50\,\text{kN}$, maximum chord force
$90\,\text{kN}$ compression, third-bay diagonal $14.14\,\text{kN}$
tension, and central vertical zero. A Warren truss (alternating
diagonals, no verticals) is checked the same way against a textbook
answer; the only change is the connectivity list. A careful solver also
returns the condition number of $A$, which spikes as the truss
approaches a mechanism geometry.
\end{exercise}

\begin{exercise}
Implement a numerical beam shear-and-moment-diagram routine that
accepts a beam length, support locations and types, point loads,
distributed loads, and applied moments, and returns $V(x)$ and
$M(x)$ as discretised arrays. Verify the maximum bending moment
of a simply-supported beam under uniform line load against $w_0 L^2
/ 8$.

\paragraph{Solution sketch.} The reference implementation is
\texttt{code/shear\_moment.py}: compute the reactions from global
equilibrium, sample the distributed load on a grid, integrate $-w$ to
shear by a running sum, subtract point loads at the stations they cross,
then integrate the shear to moment by the trapezoidal rule and subtract
applied couples at their stations. The closure check is that the shear
plus the right reaction returns to zero at the far support. Verification:
a uniform $w_0$ over span $L$ returns $M_{\max} = w_0 L^2/8$ at midspan
to within the discretisation error, and the central-point-load case
returns $PL/4$. The discretised maximum approaches the closed form as the
grid refines; with two thousand points the error is under a tenth of a
percent.
\end{exercise}

\begin{exercise}
Simulate the load-versus-deflection response of the project's
wooden truss using a simple linear-elastic model. Compare the
predicted maximum load (where any member's axial stress exceeds the
material's ultimate strength) to the predicted joint-failure load
(where any joint's pin-shear or block-shear capacity is exceeded).
State which mode the simulation predicts will govern.

\paragraph{Solution sketch.} Run the truss solver of the first
simulation exercise on the project geometry to get member forces per
unit applied load; member force scales linearly with the applied load in
a linear-elastic model, so the failure load for each mode is the
capacity divided by the per-unit force. For each member compute the
member-failure load (ultimate stress times area, divided by per-unit
axial force) separately for tension members and compression members
(compression members may buckle first, $P_{cr} = \pi^2 EI/L_e^2$). For
each joint compute the joint-failure load (the smallest of bolt-shear,
bearing, net-section, and block-shear capacities, divided by the
per-unit member force into that joint). The governing load is the
smallest over all members and joints, and the governing mode is whichever
produced it. For a glued-gusset wooden truss the joint typically governs,
as the sample test data in \texttt{data/truss\_load\_test.csv} show: the
recorded failure was gusset shear-out at a lower joint, a joint mode, not
a member fracture.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a simply-supported wooden joist to span $4\,\si{\meter}$
carrying a $2.5\,\text{kN}/\text{m}$ residential live load. Specify
the joist's cross-section and identify the dominant failure mode
(bending, shear, or deflection). State the analysis and the
governing criterion.

\paragraph{Solution sketch.} The maximum moment is $w_0 L^2/8 =
2.5 \times 16 / 8 = 5\,\text{kN}\cdot\text{m}$. Required section modulus
$S = M/\sigma_{\text{allow}}$; for sawn softwood
$\sigma_{\text{allow}} \approx 8\,\text{MPa}$ gives $S =
5000/(8 \times 10^{6}) = 6.25 \times 10^{-4}\,\si{\meter\cubed}$. A
rectangle has $S = bh^2/6$, so a $50 \times 250\,\si{\milli\meter}$
joist gives $S = 0.05 \times 0.25^2/6 = 5.2 \times 10^{-4}$, marginal;
a $50 \times 300\,\si{\milli\meter}$ gives $7.5 \times 10^{-4}$, ample.
Check deflection: $\delta = 5 w_0 L^4/(384 EI)$ with $E \approx
10\,\text{GPa}$ and $I = bh^3/12$; for the $50 \times 300$ section
$I = 1.1 \times 10^{-4}\,\si{\meter\tothe{4}}$ and $\delta \approx
5 \times 2500 \times 256/(384 \times 10^{10} \times 1.1 \times 10^{-4})
\approx 8\,\si{\milli\meter}$, about $L/500$, inside the usual $L/360$
limit. Shear is rarely critical for slender wooden joists. The governing
criterion is bending strength for the section selection and deflection
as the serviceability check; specify $50 \times 300\,\si{\milli\meter}$.
\end{exercise}

\begin{exercise}
Design a steel hanger rod to suspend a $500\,\si{\kilogram}$ light
fixture from a structural beam $3\,\si{\meter}$ above the floor.
Specify the rod's diameter, the bolt and nut at each end, and the
factor of safety against tension fracture and against the beam's
own local-bearing failure at the rod's attachment.

\paragraph{Solution sketch.} The load is $W = 500 \times 9.81 \approx
4{,}900\,\text{N}$. Choose a target factor of safety of, say, four
against the rod's ultimate tensile strength
$F_u \approx 400\,\text{MPa}$, so the design stress is
$100\,\text{MPa}$ and the required area is $A = 4900/(10^{8}) =
4.9 \times 10^{-5}\,\si{\meter\squared} = 49\,\si{\milli\meter\squared}$,
diameter $d = \sqrt{4A/\pi} \approx 8\,\si{\milli\meter}$. Specify an
$\text{M10}$ rod (tensile-stress area $58\,\si{\milli\meter\squared}$,
giving margin over the threaded section). The nut bearing area on the
beam flange must keep bearing stress below the beam material's limit:
for a steel beam this is rarely critical with a standard washer; for a
timber beam add a plate washer to spread the $4.9\,\text{kN}$ over an
area large enough to keep the perpendicular-to-grain bearing stress
under a few megapascals, roughly $10^{3}\,\si{\milli\meter\squared}$, a
$35\,\si{\milli\meter}$ square washer. State both factors of safety:
four on rod tension and the bearing check governed by the beam material.
\end{exercise}

\begin{exercise}
Design a $10\,\si{\meter}$ Pratt truss to support a deck carrying
$5\,\text{kN}/\text{m}$ of uniform live load plus the truss's own
weight. Specify the truss depth, bay spacing, and the maximum
axial force in any member. State the dominant failure mode for the
critical member.

\paragraph{Solution sketch.} Choose a depth around span/10 to span/8,
say $h = 1.2\,\si{\meter}$, and six bays of $a = 1.67\,\si{\meter}$. The
distributed deck load of $5\,\text{kN}/\text{m}$ over $10\,\si{\meter}$
totals $50\,\text{kN}$ delivered as joint loads of roughly
$8.3\,\text{kN}$ at the five interior joints; add the truss self-weight,
perhaps ten percent, and round joint loads to $9\,\text{kN}$. The
maximum chord force, by the formula derived in the calculation block,
is $F_{\text{chord}} = nPL/(8h) = 6 \times 9 \times 10/(8 \times 1.2)
\approx 56\,\text{kN}$. The most-loaded compression member is the top
chord at $56\,\text{kN}$; its dominant failure mode is buckling, checked
by $P_{cr} = \pi^2 EI/L_e^2$ against the bay length, so the top chord is
sized for buckling resistance rather than crushing. The tension bottom
chord is sized for net-section tension at its bolted splices. A deeper
truss lowers the chord force at the cost of more material in the
diagonals and verticals, the design trade the depth choice resolves.
\end{exercise}

\begin{exercise}
Design a pin connection for a $25\,\si{\milli\meter}$-diameter
two-force member carrying a tensile load of $80\,\text{kN}$. Specify
the pin diameter and the connecting plates' thickness and width.
Check against bolt shear, bearing, net-section tension, and block
shear.

\paragraph{Solution sketch.} Design load $80\,\text{kN}$ in tension.
\emph{Pin shear} (double shear through two plates): required area per
shear plane $A = V/(2 \times 0.6 F_u)$; with $F_u = 800\,\text{MPa}$,
$A = 80000/(2 \times 0.6 \times 8 \times 10^{8}) = 8.3 \times 10^{-5}\,
\si{\meter\squared}$, diameter about $10\,\si{\milli\meter}$, so choose
$\text{M16}$ for margin and a factor of safety. \emph{Bearing} on the
plate: stress $= V/(d\,t)$ must stay below the plate bearing limit
(roughly $2 F_y$); for a $16\,\si{\milli\meter}$ pin and a target stress
of $500\,\text{MPa}$, $t \geq 80000/(0.016 \times 5 \times 10^{8})
= 10\,\si{\milli\meter}$. \emph{Net-section tension}: net area
$(w - d)t$ times plate $F_u$ must exceed the load with a factor of
safety; for $t = 12\,\si{\milli\meter}$ and $F_u = 400\,\text{MPa}$,
net width needed is $80000/(0.012 \times 4 \times 10^{8}) +
0.016 = 33\,\si{\milli\meter}$, so a $60\,\si{\milli\meter}$ plate width
gives ample net section. \emph{Block shear}: provide edge distance of at
least $1.5d \approx 24\,\si{\milli\meter}$ behind the pin so the
tear-out planes carry more than the load. Specify $\text{M16}$ pin,
$12\,\si{\milli\meter}$ plates, $60\,\si{\milli\meter}$ wide, with
$25\,\si{\milli\meter}$ end distance; the governing check is bearing,
which set the plate thickness.
\end{exercise}

\begin{exercise}
Design a cantilever bracket to support a hanging shop light of
mass $20\,\si{\kilogram}$ at a horizontal offset of $400\,\si{
\milli\meter}$ from a wall. Specify the bracket's section, the
attachment to the wall, and the factor of safety. State the
worst-case loading scenario you have considered.

\paragraph{Solution sketch.} The static load is
$W = 20 \times 9.81 \approx 196\,\text{N}$ at offset
$0.4\,\si{\meter}$, giving a wall moment $M = 196 \times 0.4 \approx
78\,\text{N}\cdot\text{m}$. The worst case is not the static light but a
worker leaning on or hanging from the bracket: assume an accidental
$1\,\text{kN}$ at the tip, raising the moment to $400\,\text{N}\cdot
\text{m}$. Size the bracket section for this: required section modulus
$S = M/\sigma_{\text{allow}}$; for steel at $150\,\text{MPa}$ allowable,
$S = 400/(1.5 \times 10^{8}) = 2.7 \times 10^{-6}\,\si{\meter\cubed}$, met
by a modest angle or rectangular tube. The wall attachment carries the
moment as a couple: with two bolts spaced $100\,\si{\milli\meter}$
vertically, the tension in the upper bolt is $M/\text{spacing} =
400/0.1 = 4\,\text{kN}$, which sets the bolt and anchor. Specify a
factor of safety of about three on the accidental load. The worst-case
scenario stated: a person treating the bracket as a handhold, an order
of magnitude above the design light load.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A wooden footbridge sags visibly under foot traffic. The bridge is
a simply-supported deck on two abutments, $5\,\si{\meter}$ apart,
with two longitudinal timbers under the deck. The deck is $50\,\si{
\milli\meter}$ thick. Draw the FBD of one longitudinal under live
load, identify the critical section, and state whether the dominant
deflection mode is bending or shear.

\paragraph{Solution sketch.} Model one longitudinal as a
simply-supported beam on the two abutments carrying half the deck's live
load as a uniform line load. The FBD shows the two abutment reactions
upward and the distributed foot traffic downward; the critical section
for moment is midspan, where $M = w_0 L^2/8$, and for shear at the
supports, where $V = w_0 L/2$. For a span-to-depth ratio of
$5000/50 = 100$, the beam is slender, and slender beams deflect almost
entirely in bending: the bending deflection scales as $L^4$ while the
shear deflection scales as $L^2$, so at this slenderness bending
dominates by orders of magnitude. The visible sag is a bending
deflection, and the cure is more depth (the cube law on $I$), not a
thicker deck. A rule of thumb: shear deflection matters only for stubby
beams with span-to-depth ratios below about ten.
\end{exercise}

\begin{exercise}
A bolted steel connection joining two plates in a tension link has
developed a visible crack along the line between the bolt and the
plate edge. Identify which of the four bolted-joint failure modes
the crack indicates is in progress, and what the next inspection
finding would distinguish it from the alternative.

\paragraph{Solution sketch.} A crack running from the bolt hole toward
the nearest plate edge, along the line of the load, points to
net-section tension failure: the plate is tearing across its reduced
section. The alternative reading is block shear, where the crack would
not run straight to the edge but would turn and run parallel to the load
along two shear planes behind the bolt, defining a block about to tear
out. The distinguishing inspection finding is the crack's geometry: a
single transverse crack across the net section indicates net-section
tension; a pair of cracks parallel to the load joined by a transverse
crack behind the bolt indicates block shear. A secondary tell is
deformation: bearing-driven hole elongation (the hole stretched into an
oval ahead of the bolt) would point instead toward a bearing problem
that has not yet cracked. The next step is to measure the hole and map
the crack path before deciding which capacity was exceeded.
\end{exercise}

\begin{exercise}
A roof truss in a residential attic shows a visible gap at one of
the bottom-chord joints. Identify what loading change or what
construction defect could have produced the gap, and which truss
members now carry a different force than the original design
intended.

\paragraph{Solution sketch.} A gap at a bottom-chord joint means the
joint has slipped or pulled apart, which happens when the joint's
fastening (a gusset plate, splice, or nailed connection) yields under
the tension the bottom chord carries. Candidate causes: under-nailed or
under-bolted splice, a fastener that backed out or sheared, wood
shrinkage opening a glued joint, or an overload from a heavier roof
covering than designed. Once the bottom chord splice slips, the chord can
no longer carry its full tension across the gap, so the load redistributes:
the diagonals and the top chord pick up more, and the truss begins to act
partly as a mechanism, with deflections growing. The members most
affected are the bottom-chord segments adjacent to the gap (now carrying
reduced or zero tension) and the diagonals nearest the gap (now carrying
more, as they take over the load path the chord dropped). The inspection
priority is the integrity of the remaining bottom-chord connections,
because a second slip turns a degraded truss into a collapsed one.
\end{exercise}

\begin{exercise}
A shelf bracket has rotated visibly from its original alignment.
Draw the FBD of the bracket and identify which connection (the
wall attachment or the shelf-to-bracket connection) is most likely
to have failed. State the failure mode you would inspect for first.

\paragraph{Solution sketch.} The FBD of an L-bracket loaded by a shelf
shows a downward load at the shelf's centre of gravity, offset from the
wall, which the wall attachment must resist as a couple: the upper fixing
in tension (pulling out of the wall) and the lower in compression
(bearing into the wall). A bracket that has rotated has lost the tension
fixing: the upper anchor has pulled out, letting the bracket pivot about
its lower edge. The wall attachment is the likely failure, not the
shelf-to-bracket connection, because the shelf-to-bracket joint carries
mostly shear while the wall attachment carries the full overturning
couple amplified by the short vertical spacing of its fixings. Inspect
first for upper-anchor pull-out: a plug that has worked loose in
drywall, a stripped screw, or a fixing that was never set into a stud.
The rotation is the geometric signature of a tension fixing that let go.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Re-read the NBS report on the Hyatt Regency walkway collapse \cite{
acc:nbs-hyatt-1982}. Draw the FBD of the modified upper box-beam
connection at full scale, identifying the bolted-joint failure
modes (bolt shear, bearing, net-section tension, block shear). State
which mode the NBS report identifies as the actual mechanism, and
which of the other three modes would have governed had the actual
mode not.

\paragraph{Solution sketch (rubric).} The strong response draws the
modified upper connection at full scale: the box-beam, the hanger rod
passing through it, the nut and washer bearing on the lower flange, and
the doubled load the modification imposed (the redesign put the
fourth-floor walkway's load onto the same nut that the original single
rod was to share). The FBD shows the nut-and-washer reaction bearing on
the box-beam's lower flange, and the four bolted-joint modes are checked
at that interface. The NBS report identifies the mechanism as the nut
and washer pulling through the box-beam lower flange, a local tear-out
that is the block-shear and bearing family of failure
\cite{acc:nbs-hyatt-1982}. The response should state which of the
remaining modes would have governed next had the flange held: the rod's
own net-section tension at its thread is the natural runner-up. The
load-bearing point of the exercise is that the doubled load and the
absent joint-level FBD, not any exotic mechanism, produced the collapse;
a routine connection check at stamp time would have caught it. The
grader looks for an explicit full-scale connection drawing, all four
mode capacities computed or estimated, and the correct identification of
the actual mechanism.
\end{exercise}

\begin{exercise}
A timber pedestrian bridge collapses under a stationary tour group
of $30$ people. The investigator finds that one of the bottom-chord
splice connections used four bolts instead of the eight specified
by the design. State, in $200$ to $300$ words, which joint failure
mode the under-bolted splice most likely produced, and what
calculation a stamp-time review should have done to catch the
discrepancy.

\paragraph{Solution sketch (rubric).} Thirty people at roughly
$750\,\text{N}$ each is about $22\,\text{kN}$ of live load, which the
bottom chord carries in tension and the splice must transfer across its
discontinuity. Halving the bolt count from eight to four halves the
splice's bolt-shear and bearing capacities, so a splice with a design
factor of safety near two now sits near unity and fails under a full
crowd. The likely mode is bolt shear or bearing at the splice, the two
that scale directly with bolt count; a strong answer names bolt shear as
the most probable, since four bolts in shear is the capacity most
obviously halved. The stamp-time calculation that should have caught it
is the connection capacity check: compute the splice tension from the
crowd load, compute the four-bolt shear and bearing capacities, and
confirm the smaller exceeds the demand with the required factor of
safety. The check is a five-minute calculation that the reviewer either
never did or did against the specified eight bolts rather than the four
built. The grader looks for the load estimate, the recognition that
halving bolts halves the count-dependent capacities, the correct mode,
and the specific stamp-time check named.
\end{exercise}

\begin{exercise}
A residential roof truss has been modified by removal of a
collar tie to accommodate a skylight. State the new internal force
distribution in the truss after the modification (which members now
carry forces they did not, and which carry different forces from the
original analysis) and identify the failure mode you would inspect
for during the next snow load.

\paragraph{Solution sketch (rubric).} The collar tie ties the two
rafters together near the ridge and carries tension under gravity load,
keeping the rafters from spreading. Removing it removes that tension
restraint: the rafters now push their feet outward at the wall plates,
the wall plates and the bottom chord (ceiling joist, if it is acting as
a tension tie) pick up the spreading thrust, and the rafters carry
increased bending because they have lost an intermediate support point.
Members that change: the bottom-chord tie now carries more tension (or,
if it was never detailed as a tie, the wall connections now carry an
outward thrust they were not designed for); the rafters carry more
midspan bending. The failure mode to inspect for during the next snow
load is rafter sag and wall-plate spreading, visible as the roof
ridge dropping and the walls bowing outward at the top. A strong answer
identifies the collar tie as a tension member, traces the lost tension
into rafter thrust and bottom-chord tension, and names ridge sag plus
wall spread as the inspection target.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer asks whether the equilibrium analysis of a frame
can be done by summing forces and moments on the frame as a whole
without disassembling it. Write a one-paragraph reply that takes
the question seriously, states the conditions under which the
shortcut is acceptable, and states the conditions under which it
fails.

\paragraph{Solution sketch (rubric).} The reply takes the question
seriously: treating the whole frame as one rigid body and summing forces
and moments is a valid and useful first step, and it does give the
support reactions whenever the frame is statically determinate
externally, that is, when the external reactions number three or fewer in
the plane. What the whole-frame analysis cannot give is the internal pin
forces between members, which is the whole reason a frame is harder than
a single body. Those require disassembling the frame and writing
equilibrium for each member, using Newton's third law to pair the pin
forces. The shortcut therefore is acceptable when the only quantities
wanted are the external reactions and the frame is externally
determinate; it fails the moment the engineer needs an internal member
force, a pin reaction, or the analysis of an indeterminate frame. A
strong reply states both halves: the shortcut works for external
reactions, and disassembly is required for internal forces.
\end{exercise}

\begin{exercise}
A senior colleague argues that the truss assumptions (pin joints,
axial-only loading, no member weight) are too conservative for a
modern welded steel truss and that the analysis should be a rigid
frame from the start. Write a one-paragraph reply that takes the
argument seriously and states when each model is appropriate.

\paragraph{Solution sketch (rubric).} The colleague is partly right: a
welded steel truss does have rigid joints, and a full rigid-frame
analysis captures the secondary bending the pin-jointed model ignores.
The reply grants this and then states the trade. The pin-jointed truss
model is appropriate for member sizing and overall force flow because the
axial forces dominate and the secondary bending moments at welded joints
are usually small (the so-called secondary stresses, a few percent to
perhaps fifteen percent of the axial stress for well-proportioned
members). The rigid-frame analysis is appropriate when the secondary
stresses are not small: short stocky members, large joint eccentricities,
fatigue-critical details, or cases where the secondary bending governs a
connection. The mature practice is to size with the truss model and check
the secondary stresses, escalating to a frame analysis only where they
matter. A strong reply rejects the false choice between the two models
and states the regime where each earns its keep.
\end{exercise}

\begin{exercise}
A draft design for a balcony cantilever uses a single $20\,\si{
\milli\meter}$ bolt at the wall attachment, with the bolt analysed
as a pin (no moment). Argue, in $200$ to $300$ words, whether the
analysis is acceptable, what additional check the bolt requires,
and whether the single-bolt connection is appropriate for the
loading.

\paragraph{Solution sketch (rubric).} The pin idealisation is wrong
for a cantilever, and a strong answer says so directly. A cantilever
transmits its load to the wall as a moment, not just a force, so the
attachment must resist an overturning couple. A single bolt analysed as a
pin (two force components, no moment) cannot carry that couple: a single
fastener has no moment arm against rotation, so the connection is a
mechanism that pivots about the bolt unless something else (bearing of
the bracket plate against the wall) supplies the resisting couple. The
additional check the bolt requires is its tension under the overturning
moment, computed from the moment divided by the effective lever arm
between the bolt and the bracket's bearing edge, plus the direct shear
from the vertical load, combined. The single-bolt connection is
inappropriate for a balcony cantilever: it provides no redundancy, and a
single bolt failure drops the balcony. The standard detail uses at least
two bolts vertically separated so the couple is carried as a bolt-tension
pair. A strong reply names the moment the pin model hides, the
tension-plus-shear check, and the inadequacy of a single fastener for a
life-safety cantilever.
\end{exercise}

\begin{exercise}
Identify a built structure in your immediate environment whose
internal load path you have never traced explicitly. State, in
$200$ to $300$ words, the chosen body, the support type at each
end, the dominant internal force (axial, shear, or moment) at the
critical section, and the failure mode you would inspect for first.

\paragraph{Solution sketch (rubric).} The exercise is open by design;
the rubric grades the discipline of the trace, not the structure chosen.
A strong answer picks a concrete object (a shelf, a doorway lintel, a
balcony rail, a bike frame's top tube), states the support at each end in
the chapter's vocabulary (pin, roller, fixed, cable), identifies the
dominant internal force at the critical section with a reason (a lintel
over an opening carries bending, peaking at midspan; a hanging shelf rod
carries axial tension; a railing post carries bending from a lateral
push at its base), and names the failure mode that the dominant force
drives (bending fracture or excessive deflection for the lintel,
fastener pull-out for the shelf, base-connection failure for the post).
The grader looks for a coherent FBD-to-failure chain: support types,
load path, dominant internal force, and a failure mode that follows from
that force rather than a generic worry. The point of the exercise is to
turn the chapter's machinery on the reader's own surroundings.
\end{exercise}
