\chapter{Composites}
\label{vol05:ch08}

\epigraph{The properties of a composite are not the properties of its
constituents. They are the properties of the interface. A glass fibre
in a polymer matrix is a stronger material than either; a glass fibre
in poor contact with the matrix is a weaker material than the matrix
alone.}{Adapted from K.~K. Chawla, working scope of composite
materials engineering \cite[ch.~1, pp.~3--7]{text:chawla-composites}.}

\chapmeta{Half-life: durable. Archetypes: scaling, optimisation. Exercise target: 24. Project track: physical fabrication and load test. Hazard class: medium (epoxy resin curing, glass-fibre handling, sharp cut edges; respirator and gloves for hand lay-up). Reader path default: standard, with sections explicitly tagged. Named cases: Columbia leading-edge RCC failure (2003) referenced for composite-laminate impact failure; described by mechanism only for in-service composite components.}

\noindent
A composite is two or more materials combined to produce properties no
single constituent possesses. The chapter develops the working
composite classes (fibre-reinforced, particulate, laminate, sandwich),
the rule-of-mixtures arithmetic that predicts longitudinal and
transverse properties, the anisotropy that follows from fibre
orientation, the three reinforcement fibres that carry most of the
engineering load (glass, carbon, aramid), the four manufacturing
processes that turn fibre and resin into a part, the sandwich-panel
geometry that buys bending stiffness at low areal mass, and the three
brittle-composite failure modes the engineer must screen for.
Composites do not replace metals; they sit beside metals as the
materials of choice when the load path is well understood and the
specific stiffness or specific strength is the binding constraint.

\section{Composite types: fibre-reinforced, particulate, laminate}\pathtag{core}

A \engterm{composite material} is a combination of two or more
distinct constituents at a length scale below that of the engineered
part but above that of the atomic lattice. The constituents retain
their identities (a metal alloy is not a composite by this definition,
because the solute atoms are dissolved in the lattice). The working
classification is by reinforcement geometry and by matrix family.

\engterm{Fibre-reinforced composites}. The reinforcement is long
fibres ($L/d \gg 100$, typically $L/d > 10{,}000$ for continuous fibres,
$L/d \sim 10$--$100$ for chopped fibres). The fibres carry the load;
the matrix holds them in place, transfers load between them, and
protects them. Working examples: glass-fibre-reinforced polyester
(GFRP) for boat hulls, wind-turbine blades, pipe linings, automotive
body panels; carbon-fibre-reinforced epoxy (CFRP) for aerospace
structure, premium sporting goods, and an increasing share of premium
automotive body panels; aramid-fibre-reinforced epoxy for ballistic
armour, brake-pad linings, and tyre cords.

\engterm{Particulate composites}. The reinforcement is short, equiaxed
inclusions: ceramic particles in a metal matrix (tungsten-carbide in
cobalt for cutting tools, silicon-carbide in aluminium for engine
components), rubber particles in a brittle polymer matrix (ABS, HIPS),
or aggregate in cement (concrete, which is a particulate composite of
gravel and sand in a cement-paste matrix). The properties are
approximately isotropic; the rule of mixtures applies with averaged
constants.

\engterm{Laminate composites}. The composite is a stack of plies, each
ply itself a fibre-reinforced sheet. The ply directions are chosen to
match the load path: a wing skin under bending might use a $0^\circ /
\pm 45^\circ / 90^\circ$ stack to carry the bending, the shear, and the
hoop loads with separate ply families. The full laminate is anisotropic
in a structured way; classical lamination theory (section~8.3) gives
the working machinery for computing the laminate's stiffness from the
ply properties and the lay-up sequence.

\engterm{Sandwich structures}. Two stiff skins separated by a light
core form a composite optimised for bending. The skins carry tension
and compression; the core carries shear and holds the skins apart. The
working core materials are honeycomb (aluminium, Nomex, polypropylene),
polymer foam (PVC, polyurethane, PMI), and balsa wood. The skins are
fibre-reinforced composites or thin metal sheets. Sandwich panels are
the workhorses of aircraft interior panels, premium yacht hulls, and
modern shipping containers. Section~8.6 develops the sandwich-panel
mechanics; section~10 of this volume returns to the sandwich panel as
an Ashby-style worked example.

\engterm{Hybrid composites}. The reinforcement combines two or more
fibre types (carbon plus glass, carbon plus aramid). Hybrids trade off
the brittleness of high-modulus fibres against the toughness of
lower-modulus fibres, and trade the higher cost of carbon against the
lower cost of glass. The aerospace literature uses hybrids
extensively; the working selection logic returns in section~10.

\engterm{Functionally graded composites}. The volume fraction or the
constituent identity varies continuously through the part: a thermal
barrier coating that transitions from metallic substrate through a
graded layer to a ceramic outer surface; a turbine blade whose root is
metallic and whose airfoil tip is ceramic; a bone-replacement implant
whose density varies to match the bone's density gradient. Functionally
graded materials are an active research area as of 2025; production
parts in commercial service remain a small fraction of the composite
market.

\section{Rule of mixtures}\pathtag{core}

The simplest model of a composite property is the volume-fraction-
weighted average of the constituent properties. For a composite with
fibre volume fraction $V_f$ and matrix volume fraction $V_m = 1 - V_f$,
the longitudinal modulus is
\[
E_1 = V_f E_f + V_m E_m \quad (\text{Voigt average, equal-strain
assumption}).
\]
The relation follows from the assumption that both constituents
experience the same longitudinal strain (good for continuous fibres
aligned with the loading direction). The transverse modulus is bounded
below by the Reuss (equal-stress) average:
\[
\frac{1}{E_2} = \frac{V_f}{E_f} + \frac{V_m}{E_m} \quad
(\text{Reuss average, equal-stress assumption}).
\]
The two averages bracket the true property. For continuous-fibre
composites the longitudinal modulus is well predicted by the Voigt
average; the transverse modulus lies between the two bounds and is
usually estimated by the empirical Halpin-Tsai equation,
\[
\frac{E_2}{E_m} = \frac{1 + \xi \eta V_f}{1 - \eta V_f}, \quad
\eta = \frac{(E_f/E_m) - 1}{(E_f/E_m) + \xi},
\]
with $\xi \approx 2$ for the transverse modulus of a unidirectional
laminate. The Halpin-Tsai expression interpolates between Voigt
($\xi \to \infty$) and Reuss ($\xi = 0$) and matches experimental
transverse moduli to about $\pm 5\,\%$ \cite[ch.~6]{
text:chawla-composites}.

\engterm{Working numbers for unidirectional composites}. A typical
high-strength carbon-fibre/epoxy laminate at $V_f = 0.6$ has
$E_1 \approx 140\,\text{GPa}$, $E_2 \approx 10\,\text{GPa}$, in-plane
shear modulus $G_{12} \approx 5\,\text{GPa}$, and Poisson ratio $\nu_{12}
\approx 0.3$. The longitudinal-to-transverse modulus ratio of $\sim 14$
is the anisotropy that drives lay-up design: a single-ply structure is
$14$ times stiffer along the fibre than across it. The same composite
in cross-ply $(0/90)_s$ stacking averages to $E \approx 75\,\text{GPa}$
in both directions but loses the maximum stiffness available from
unidirectional plies.

\engterm{Strength rule of mixtures}. The longitudinal tensile strength
of a unidirectional composite is governed by the fibre breaking
strain. If the fibres are the brittle constituent (fibre breaking
strain $\varepsilon_f^*$ less than matrix breaking strain
$\varepsilon_m^*$), the composite fails at $\sigma_c =
V_f \sigma_f^* + V_m E_m \varepsilon_f^*$ when the fibres break and
the matrix is loaded to the fibre breaking strain. For $V_f > V_f^{\min}
\approx \sigma_f^*/(\sigma_m^* + V_f \sigma_f^*)$, the composite carries
load until the fibres break; below $V_f^{\min}$ the matrix carries
more load than the fibre and the composite strength does not benefit
from adding fibres. The working $V_f$ for engineering composites is
$0.4$ to $0.65$; lower volume fractions fall below the threshold,
higher volume fractions leave too little matrix to wet the fibres.

\engterm{The role of the interface}. The fibre-matrix interface is the
load-transfer mechanism and the dominant failure-initiation site. A
strong interface gives a stiff, high-strength composite that fails in
a brittle, catastrophic mode (the crack runs across the interfaces
because the interface is stronger than the matrix); a weak interface
gives a tougher composite that fails by progressive
debonding-and-pullout (the crack deflects along the interface and is
arrested by friction at the bridging fibres). The fibre coating,
called a \engterm{sizing} for glass fibres or a \engterm{surface
treatment} for carbon fibres, is the engineered chemistry that sets
the interface strength. Volume V Chapter 11 returns to manufacturing-
process control of the interface.

\begin{principle}[The rule of mixtures predicts longitudinal stiffness;
the matrix and the interface predict everything else]
The Voigt average $E_1 = V_f E_f + V_m E_m$ gives the longitudinal
modulus of a unidirectional composite to within a few percent. The
transverse modulus, the in-plane shear modulus, the through-thickness
strength, the impact resistance, and the failure mode are all set by
the matrix and by the fibre-matrix interface, not by the fibre. A
composite design that optimises only the fibre fraction will produce a
stiff part that breaks at the matrix or the interface, not at the
fibre.
\end{principle}

\section{Anisotropy and lay-up design}\pathtag{standard}

A unidirectional ply has different elastic constants along, across,
and at $45^\circ$ to the fibre direction. The full in-plane elastic
behaviour of a ply is captured by the four constants $E_1$, $E_2$,
$G_{12}$, and $\nu_{12}$ (with $\nu_{21}$ following from the
reciprocity relation $\nu_{12}/E_1 = \nu_{21}/E_2$). The ply's
compliance matrix in the ply principal axes is
\[
\mathbf{S} = \begin{bmatrix}
1/E_1 & -\nu_{12}/E_1 & 0 \\
-\nu_{12}/E_1 & 1/E_2 & 0 \\
0 & 0 & 1/G_{12}
\end{bmatrix},
\]
and the ply stiffness matrix $\mathbf{Q} = \mathbf{S}^{-1}$.

\engterm{Off-axis loading}. A ply oriented at angle $\theta$ to the
loading direction has transformed stiffness
\[
\bar{\mathbf{Q}}(\theta) = \mathbf{T}^{-1}(\theta) \mathbf{Q}
\mathbf{T}^{-T}(\theta),
\]
with $\mathbf{T}(\theta)$ the standard rotation matrix for second-
order tensors in two dimensions. The transformed compliances admit a
working expansion in $\theta$ that the designer learns by
practice: the modulus along the fibre falls from $E_1$ at $\theta = 0$
to $E_2$ at $\theta = 90^\circ$, passing through a minimum of
$\sim E_2 / (1 - 2 \nu_{12} \sqrt{E_2/E_1})$ near $45^\circ$. The
$45^\circ$ direction is the most flexible direction in a unidirectional
ply, which is why structural laminates rarely use a single-direction
stack.

\engterm{Classical lamination theory}. A laminate is a stack of plies
with chosen orientations $\theta_k$ and thicknesses $t_k$. The
laminate's in-plane stiffness matrix $\mathbf{A}$ is the sum of the
transformed ply stiffnesses weighted by ply thickness:
\[
\mathbf{A} = \sum_k \bar{\mathbf{Q}}_k t_k.
\]
The bending stiffness $\mathbf{D}$ is the second moment of the ply
stiffnesses about the laminate midplane, and the coupling stiffness
$\mathbf{B}$ is the first moment. A symmetric laminate (ply
orientations symmetric about the midplane) has $\mathbf{B} = 0$:
in-plane and bending loads decouple, and the laminate does not warp
under temperature change. A balanced laminate (equal numbers of
$+\theta$ and $-\theta$ plies) has zero in-plane shear-extension
coupling. Most engineering laminates are designed symmetric and
balanced.

\engterm{Lay-up nomenclature}. A symmetric laminate is written as
$[\theta_1/\theta_2/\dots/\theta_n]_s$, indicating that the
half-laminate $\theta_1, \theta_2, \dots, \theta_n$ is mirrored at the
midplane. $[0/45/-45/90]_s$ is a quasi-isotropic eight-ply laminate
with one-fourth of the plies in each of four directions. The
quasi-isotropic lay-up has the same in-plane stiffness in every
direction, equal to approximately $3/8 (E_1 + E_2)$ for a ply with
$E_1 \gg E_2$. Quasi-isotropic lay-ups are the working baseline for
aerospace skin panels where the load directions are not known in
advance.

\engterm{The ten-percent rule}. Aerospace lay-up practice keeps at
least $10\,\%$ of plies in each of four orientations ($0^\circ$,
$+45^\circ$, $-45^\circ$, $90^\circ$). The rule prevents catastrophic
loss of stiffness in any direction under off-axis loading, secondary
load paths, manufacturing-misalignment errors, and damage-tolerance
considerations. The rule is published as a design guideline in the
working aerospace structural-design manuals and is the canonical
engineering expression of the principle that lay-up should not be
optimised at the margin against any specific load case.

\engterm{Optimisation of lay-ups}. The orientation of every ply is a
discrete design variable (orientation $\theta$ from a chosen set,
typically $\{0, 15, 30, 45, 60, 75, 90\}^\circ$). The lay-up optimiser
chooses the orientations to maximise stiffness or strength under given
load cases. The problem is a discrete nonlinear programme; the working
solver is a genetic algorithm or simulated annealing for industrial
parts, and a continuous parametric sweep for student exercises.

\section{Carbon fibre, glass fibre, aramid}\pathtag{standard}

Three reinforcement fibres carry most of the engineering composite
volume.

\engterm{Glass fibres}. Pulled from a melt of soda-lime-borosilicate-
alumina glass (E-glass) or a stronger calcium-aluminoborosilicate
(S-glass), at diameters of $10$--$25\,\si{\micro\meter}$. E-glass has
$E = 70\,\text{GPa}$, $\sigma^* \approx 3.4\,\text{GPa}$ at the
working fibre length, $\varepsilon^* \approx 4.8\,\%$. S-glass has
$E = 89\,\text{GPa}$, $\sigma^* \approx 4.6\,\text{GPa}$,
$\varepsilon^* \approx 5.7\,\%$. Density $\rho \approx 2.5\,
\text{g}/\text{cm}^3$. Cost (as of 2024) $\sim \$1$--$2$ per
kilogram for E-glass roving, $\sim \$10$--$15$ per kilogram for
S-glass. The bulk of fibreglass-reinforced plastic is E-glass; the
premium aerospace and ballistic-protection applications use S-glass.

\engterm{Carbon fibres}. Produced by heat-treating a precursor fibre
(polyacrylonitrile, PAN, the working precursor for the bulk of
production; pitch for the highest-modulus grades) under tension at
temperatures up to $\sim 2500\,\degC$. The resulting fibre is
$\sim 90$--$98\,\%$ carbon arranged in graphitic ribbons aligned with
the fibre axis; the higher the heat-treatment temperature, the more
graphitic and the stiffer the fibre. Working ranges: standard-modulus
(SM) fibres at $E \approx 230\,\text{GPa}$, $\sigma^* \approx 3.5\,
\text{GPa}$, $\varepsilon^* \approx 1.5\,\%$; intermediate-modulus
(IM) fibres at $E \approx 290\,\text{GPa}$, $\sigma^* \approx 5.5\,
\text{GPa}$, $\varepsilon^* \approx 1.9\,\%$; high-modulus (HM) fibres
at $E \approx 500\,\text{GPa}$ but with reduced strength
($\sigma^* \approx 3.0\,\text{GPa}$) and strain. Density $\rho \approx
1.8$ -- $1.9\,\text{g}/\text{cm}^3$. Cost: $\sim \$15$--$40$ per
kilogram for IM/aerospace grades as of 2024, with substantial
volatility tied to the cost of the PAN precursor and the energy of
the carbonisation kiln. The cost penalty over glass is the working
constraint on penetration into mass-market applications.

\engterm{Aramid fibres}. Aromatic-polyamide fibres (Kevlar, Twaron) at
$E \approx 70$--$130\,\text{GPa}$, $\sigma^* \approx 3.0$--$3.6\,
\text{GPa}$, $\varepsilon^* \approx 2.5\,\%$ -- $4.0\,\%$, density
$\rho \approx 1.45\,\text{g}/\text{cm}^3$. Aramid fibres have the
highest specific tensile strength of the three (strength-to-density)
but lower compressive strength than glass or carbon (aramid fibres
microbuckle under compression, an artefact of their long, oriented-
molecule structure). The working applications are ballistic armour
(where tension and toughness matter, compression does not), tyre
cords, brake-pad linings, and as a hybrid component with carbon in
sandwich panels.

\engterm{Other reinforcement fibres}. Polyethylene (UHMWPE: Dyneema,
Spectra) at higher specific strength than aramid, with the trade-off
of creep at room temperature. Ceramic fibres (silicon carbide, alumina,
mullite) for high-temperature composites. Natural fibres (jute, flax,
hemp) for low-cost panels in interior automotive trim and consumer
goods. Boron fibres (CVD-deposited on a tungsten core) for niche
high-modulus and high-temperature applications.

\engterm{Comparison on specific properties}. The engineering comparison
is by specific stiffness ($E/\rho$) and specific strength
($\sigma^*/\rho$) rather than by absolute values, because the working
composite carries its own weight. Carbon-fibre IM at $E/\rho \approx
160\,\text{MN}\cdot\text{m}/\text{kg}$ exceeds aluminium (26),
titanium (27), and high-strength steel (26) by a factor of six. The
specific strength of aramid (around $2400\,\text{kJ}/\text{kg}$) is
comparable to carbon at the fibre level but in the composite is lower
because of the reduced compressive performance.

\section{Manufacturing: hand lay-up, autoclave, RTM, automated fibre placement}\pathtag{standard}

The same fibre and resin can be turned into parts of widely different
quality by different processes. The working processes are listed in
roughly increasing order of capital investment and decreasing labour
content per part.

\engterm{Hand lay-up}. Reinforcement (dry cloth or pre-impregnated
``prepreg'' plies) is placed on a mould by hand; resin is applied with
a brush or roller; air is consolidated out with hand rollers. The
process is widely used for prototype work, boat hulls, swimming pools,
and one-off architectural parts. Typical fibre volume fraction $V_f
\approx 0.3$--$0.4$; porosity $1$--$5\,\%$; substantial part-to-part
variability. Capital cost: a mould and a workbench. Cycle time: hours
to days per part. The chapter project relies on this process.

\engterm{Spray-up}. Chopped fibres and resin are sprayed onto a mould
through a chopper gun. Properties are isotropic and intermediate
between the rule-of-mixtures bounds for chopped fibres. The process is
the highest-volume composite-manufacturing process by part count;
applications include automotive body panels, bathtubs, and shower
stalls.

\engterm{Vacuum bagging}. A hand lay-up is covered with a flexible
film and the air is evacuated. Atmospheric pressure ($\sim 100\,
\text{kPa}$) consolidates the laminate, reduces porosity, and
improves fibre volume fraction to $V_f \approx 0.5$. The process is
the workhorse of small-volume aerospace and high-end marine work.

\engterm{Autoclave processing}. The vacuum-bagged lay-up is placed in
a pressure vessel that applies external pressure (typically $0.5$--$1\,
\text{MPa}$, with some military processes at $2\,\text{MPa}$) and
heated to the resin cure temperature ($120$--$180\,\degC$ for
epoxies). The combined pressure and temperature give high fibre
volume fraction ($V_f \approx 0.55$--$0.65$), porosity $< 1\,\%$, and
the working strength and stiffness of the laminate. Autoclave
processing is the standard for aerospace primary structure; the
capital cost (a large pressure vessel) and the operating cost (the
autoclave cycle is hours, and the autoclave is the throughput
bottleneck) drive design toward as few parts as possible.

\engterm{Resin transfer moulding (RTM)}. Dry preform is placed in a
closed mould; resin is injected under pressure. The process gives
high-quality parts at higher throughput than autoclave, with the
trade-off that the mould is a precision-machined closed tool. Vacuum-
assisted RTM (VARTM) drops the resin-injection pressure and uses
vacuum to drive the resin into the preform; the trade-off is somewhat
lower fibre fraction. RTM and VARTM are the working processes for
mid-volume aerospace components (Airbus A350 fuselage frames are
RTM-moulded) and for high-performance automotive body panels.

\engterm{Automated fibre placement (AFP) and automated tape laying
(ATL)}. A robotic head places narrow tapes or tows of prepreg onto a
mould, with the orientation and position computer-controlled. The
process eliminates the labour content of hand lay-up and removes the
ply-to-ply alignment variability that hand work introduces. AFP and
ATL are the working processes for large aerospace primary structure
(Boeing 787 fuselage barrels, Airbus A350 wings). The capital cost is
$\sim \$5$--$20\,\text{M}$ per machine; the cycle time is hours per
part; the working part-quality is the same as autoclave with reduced
labour.

\engterm{Pultrusion}. Continuous fibre rovings are pulled through a
resin bath and a heated shaped die to produce constant-cross-section
parts (rod, tube, angle, channel). The process gives high $V_f$ and
high strength along the pull direction, with the trade-off of
constant cross-section and continuous-length parts. Working
applications: ladder rails, structural composite reinforcing bar
(rebar), structural pipe.

\engterm{Filament winding}. A continuous tow is wound onto a rotating
mandrel at a controlled wind angle; the mandrel is then removed or
left in place. Working applications: pressure vessels (composite
overwrapped pressure vessels, COPV; the propellant tanks of essentially
every spacecraft); pipes; rocket-motor cases. The wind angle is chosen
to match the load: tension biased toward the $54.7^\circ$ optimum for
isotropic pressure loading.

\engterm{Resin systems}. The matrix is chosen to match the application:
unsaturated polyester for general-purpose marine and automotive work
(low cost, fast cure, moderate temperature and chemical resistance);
vinyl ester for chemical-resistance and corrosion-protection (higher
cost, better in marine environments); epoxy for structural aerospace
and high-performance automotive (highest cost, highest strength,
slowest cure); bismaleimide and polyimide for higher-temperature
applications ($200$--$300\,\degC$ continuous use); cyanate ester for
microwave-transparent applications. Section~11 of this volume returns
to the cure-related processing effects.

\section{Sandwich structures}\pathtag{standard}

A \engterm{sandwich panel} is a stiff laminate of two thin face sheets
separated by a thick light core. The face sheets carry tension and
compression from bending; the core carries shear and holds the faces
apart. The geometry is dimensionally analogous to an I-beam, with the
faces playing the role of the flanges and the core the role of the
web. For a panel of total thickness $h$, face-sheet thickness $t_f
\ll h$, face-sheet modulus $E_f$, and core modulus $E_c$, the bending
stiffness per unit width is approximately
\[
D = \frac{E_f t_f h^2}{2} + \frac{E_c h^3}{12} \approx
\frac{E_f t_f h^2}{2},
\]
the second term being small when $E_c \ll E_f$. Bending stiffness
scales with $t_f h^2$: doubling the core thickness doubles the panel
mass-per-unit-area only modestly (because the core is light) but
roughly quadruples the bending stiffness. The Gibson-Ashby analysis
of cellular solids gives the working scaling laws for foam and
honeycomb cores \cite[ch.~4]{text:gibson-ashby-cellular}.

\engterm{Core failure modes}. Sandwich panels fail in five working
modes that the designer must screen for:
\begin{itemize}[leftmargin=*]
\item \engterm{Face yielding or fracture}: stress in the face sheet
  exceeds its tensile or compressive strength. Sets the working span
  for given loads.
\item \engterm{Core shear failure}: shear stress in the core exceeds
  the core's shear strength. The shear flow in a sandwich is
  $\tau = V / (b h)$ approximately, with $V$ the transverse shear
  force and $b$ the panel width.
\item \engterm{Face wrinkling}: local elastic instability of a face
  sheet supported on an elastic core. The critical face stress is
  approximately $\sigma_{w} \approx 0.5 (E_f E_c G_c)^{1/3}$ for a
  panel with an isotropic core.
\item \engterm{Core indentation under concentrated load}: local
  crushing of the core under a point load that exceeds the core's
  compressive strength.
\item \engterm{Face-core debonding}: failure of the adhesive bond
  between the face sheet and the core, often initiated by impact or
  by moisture ingress.
\end{itemize}

The five modes have intersecting design constraints; the working
sandwich-panel design balances face thickness, core thickness, and
core density against the binding mode at the chosen span and load.

\engterm{Core materials}. Aluminium honeycomb at densities $30$--$200\,
\text{kg}/\text{m}^3$ and cell sizes $3$--$10\,\text{mm}$ is the
workhorse for aerospace interior panels and high-performance composite
sandwich (Boeing 787 floor panels, helicopter rotor blade cores).
Nomex (aramid-paper) honeycomb at $30$--$100\,\text{kg}/\text{m}^3$
is the workhorse for aerospace fire-resistant applications and most
commercial aerospace interior structure. Polymer foams (PVC, PMI,
polyurethane) at $40$--$200\,\text{kg}/\text{m}^3$ are the workhorse
for marine and wind-turbine applications. Balsa wood at $\rho \approx
150\,\text{kg}/\text{m}^3$ is a natural-fibre sandwich core that
outperforms many engineered foams in compression and is used in
performance boat hulls and in some wind-turbine blade roots.

\section{Failure: delamination, impact damage, fatigue in composites}\pathtag{core}

Composites fail in modes that monolithic materials do not. The chapter's
failure section names three modes that the engineer must screen for in
every composite-design review.

\engterm{Delamination}. A composite laminate is held together through
the thickness by the matrix alone; there are no fibres in the through-
thickness direction. A through-thickness tensile or shear load can
separate the plies along the matrix-rich interlaminar resin layer. The
working delamination toughness is measured by the mode I and mode II
strain-energy release rates, $G_{Ic}$ and $G_{IIc}$; for carbon-epoxy
laminates $G_{Ic} \approx 200$--$300\,\text{J}/\text{m}^2$ and
$G_{IIc} \approx 600$--$1000\,\text{J}/\text{m}^2$. These values are
two orders of magnitude below the in-plane fracture toughness; the
through-thickness load is the weakest direction.

Delamination initiates at free edges (where the stress state has a
through-thickness component because of ply mismatch in Poisson
contraction), at ply drops (where the laminate thickness changes), at
holes (where stress concentrations include through-thickness shear),
and at impact sites. The working design responses are: avoid free
edges in highly stressed regions, taper laminate thickness gradually
(rule of thumb: at least $20$ ply lengths per ply drop), avoid sharp
internal corners, design to limit through-thickness loads, and
through-thickness stitching or z-pinning for damage-tolerant designs.

\engterm{Impact damage and barely-visible impact damage (BVID)}. A
foreign-object impact on a composite laminate produces damage with a
characteristic structure: a localised face dent on the impact side, a
larger internal delamination at the back face, and a network of matrix
cracks. The damage on the impact side may be barely visible to the
unaided eye while the internal delamination is large. BVID is the
canonical aerospace-composite damage threat: a tool drop, a hailstone,
a ground-handling collision. The damage reduces the compressive
strength of the laminate by $30$--$60\,\%$ at the damage site.

The 2003 Columbia accident's left-wing leading edge failure was an
impact-damage failure of a reinforced-carbon-carbon panel
\cite{acc:caib-columbia}. A briefcase-sized piece of insulating foam
from the external tank impacted the leading-edge panel during ascent
at relative velocity $\sim 240\,\text{m}/\text{s}$. The impact created
a hole in the panel's outer mould line; on re-entry, hot plasma flowed
through the hole and burned through the wing's internal structure. The
canonical engineering lesson is that an impact whose visible damage
seemed to be limited cosmetic damage on the outer mould line in fact
penetrated the panel; the standard pre-Columbia hazard analysis had
treated the foam impact as a tile-damage threat, not a leading-edge-
panel threat. The post-Columbia inspection regime added on-orbit
imaging of the leading-edge panels before re-entry.

The engineering response to BVID is layered: damage-tolerance
allowables (work the part at stresses below the predicted residual
strength after barely-visible impact), inspection regimes (visual,
tap-test, ultrasonic, thermographic), and design choices that limit
the consequence of damage (redundant load paths, sacrificial outer
plies).

\engterm{Fatigue in composites}. Composite-laminate fatigue is
dominated by the matrix, not by the fibres. A $0^\circ$-loaded
unidirectional carbon-epoxy laminate has essentially flat S-N
behaviour: the fatigue strength at $10^7$ cycles is $\sim 80$--$90\,\%$
of the static strength. Cross-ply and quasi-isotropic laminates show
matrix cracking that accumulates over the first $10^4$ -- $10^6$ cycles;
the matrix cracks initiate delamination, which dominates the later
life. The fatigue ratio (fatigue strength to static strength at $10^7$
cycles) is $\sim 0.5$--$0.7$ for carbon-epoxy laminates and $\sim
0.3$--$0.5$ for glass-epoxy laminates. The glass-epoxy fatigue ratio is
the binding constraint on wind-turbine blade design at $20$-year
service intervals and $\sim 10^9$ -- $10^{10}$ cycles.

\engterm{Environmental degradation}. Polymer matrices absorb water
from humid air and from immersion. Water plasticises the matrix, drops
its glass transition temperature, and reduces the matrix-dominated
properties (transverse strength, in-plane shear, compression). The
working aerospace design accommodates $\sim 1\,\%$ moisture pickup in
service and applies ``hot-wet'' allowables to the design ($82\,\degC$
and $1\,\%$ moisture). UV exposure degrades the matrix surface (matrix
embrittlement, microcracking); the working response is a UV-resistant
gel coat or paint over the structural laminate.

\begin{principle}[A composite carries load along the fibres and fails
across them]
The in-plane fibre-direction properties of a composite are predicted
by the rule of mixtures and are usually excellent. The through-
thickness, transverse, and shear properties are governed by the
matrix and the interface and are usually much weaker. Every composite
failure analysis answers a single question first: was the load path
in the direction of the fibres, and if so, what off-axis load
appeared at the interface, the free edge, or the through-thickness
direction?
\end{principle}

\begin{archetype}[Scaling and optimisation]
A composite is the material class in which the designer's degrees of
freedom (fibre type, volume fraction, ply orientation, ply count,
core thickness, core density) exceed the constraints, so the design
problem is genuinely an optimisation. The chapter's working examples
all reduce to the same template: name the load cases, write the
laminate's $\mathbf{A}$ and $\mathbf{D}$ matrices, search the
discrete design space of orientations, check against the failure modes
of section~8.7, and accept the optimum. Section~10 of this volume
returns to the Ashby chart that frames the problem at the materials
level; this chapter gives the in-class search.
\end{archetype}

\begin{estimation}
\textbf{A small wind-turbine blade of length $L = 10\,\text{m}$ is to
be built from glass-epoxy laminate. Estimate the root-section laminate
thickness required to carry the bending load at the design wind
speed.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The blade is loaded by aerodynamic lift along its span. The lift force
per unit span at design conditions is approximately
$f \approx \tfrac{1}{2} \rho_{\text{air}} V^2 c C_L$, with $\rho_{\text{air}}
\approx 1.2\,\text{kg}/\text{m}^3$, design wind speed at the tip
$V \approx 60\,\text{m}/\text{s}$ (a tip-speed ratio of $7$ at a hub
wind speed of $8.5\,\text{m}/\text{s}$, both modest), chord $c \approx
0.5\,\text{m}$ at the working section, and $C_L \approx 1.0$ at the
working angle of attack. The lift per unit span is
$f \approx 1080\,\text{N}/\text{m}$.

The root bending moment is approximately
$M \approx f L^2 / 2 \approx 5.4 \times 10^4\,\text{N}\cdot\text{m}$
(using $L^2/2$ for a uniformly distributed load).

The blade root is approximately circular, with outer diameter $D \approx
0.4\,\text{m}$ and laminate thickness $t$. The section modulus is
approximately $\pi D^2 t / 4$. The peak bending stress in the laminate
is $\sigma = M / (\pi D^2 t / 4)$. For a working allowable bending
stress of $200\,\text{MPa}$ (glass-epoxy fatigue strength at $10^9$
cycles, with a safety factor), the laminate thickness is
\[
t = \frac{4 M}{\pi D^2 \sigma_{\text{allow}}} =
\frac{4 \times 5.4 \times 10^4}{\pi \times 0.16 \times 2 \times 10^8}
\approx 2.1 \times 10^{-3}\,\text{m} = 2.1\,\text{mm}.
\]

A working blade root laminate is in fact $\sim 30$--$80\,\text{mm}$
thick, an order of magnitude more than our estimate. The discrepancy
comes from extreme loads (gust loads can exceed design lift by a factor
of three to five), from buckling stability of the thin shell (which
binds well before the tensile stress allowables), and from the use of
local thickness build-ups at the root to transfer the load into the
hub. The estimate gets the order of magnitude of the bending stress
right ($O(100\,\text{MPa})$) and reveals that buckling, not tension,
binds the design.

The postmortem: the back-of-envelope captures the static-bending order
of magnitude. The design is binding on a different failure mode
(buckling) that we did not model. The estimate's value is to expose
which physical mechanism is, in fact, doing the binding.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Lay up, cure, and load-test a small fibreglass beam]
The reader fabricates a small glass-epoxy laminate beam by hand lay-up
and vacuum bagging, then load-tests it to failure in three-point
bending and identifies the failure mode by fractographic inspection.
The deliverable is a fifteen- to twenty-five-page report.

\textbf{Track}. Physical fabrication plus measurement.
\textbf{Hazard class}: medium. The epoxy is a sensitiser; gloves,
respirator with organic-vapour cartridge, eye protection, and good
ventilation are required. Glass fibre causes skin irritation; handle
in long sleeves. Cure exotherm can be substantial; size the mix and
the layup to keep the peak below the safe-handling limit.

\textbf{Materials}.

\begin{itemize}[leftmargin=*]
\item Glass-fibre cloth (woven roving or biaxial stitched, $\sim 300\,
  \text{g}/\text{m}^2$ areal mass); enough for two beams of $300 \times
  50 \times 5\,\text{mm}$.
\item Two-part epoxy resin (laminating grade) with a slow hardener and
  pot life of at least $30\,\text{min}$ at $20\,\degC$.
\item Vacuum-bag film, breather, peel-ply, sealant tape, and a small
  vacuum pump with a gauge.
\item Caul plate or flat mould (waxed glass or a melamine-coated MDF
  board).
\item Three-point bending fixture with $250\,\text{mm}$ support span,
  loading rod, and force measurement (a kitchen scale with a screw-
  loaded jig, or a laboratory force gauge).
\item Calipers and a notebook.
\end{itemize}

\textbf{Procedure}.

\begin{itemize}[leftmargin=*]
\item Cut $N$ plies of glass cloth, all to the same dimensions,
  $\sim 350 \times 60\,\text{mm}$. Decide on a lay-up: an
  $[0]_N$ unidirectional or a $[0/90]_{N/2}$ cross-ply.
\item Stack plies on the mould, applying resin to each ply with a
  brush or roller until the laminate is fully wetted. Plies should
  not show dry spots; the laminate should not be flooded with excess
  resin.
\item Cover with peel-ply, breather, and vacuum-bag film. Pull vacuum
  to $-90\,\text{kPa}$ gauge or better. Maintain vacuum for the full
  cure time (typically $16$--$24\,\text{h}$ at $20\,\degC$, or follow
  the resin's published schedule).
\item Demould; trim to final width $50\,\text{mm}$ and length
  $300\,\text{mm}$.
\item Measure and record the actual cross-section dimensions at five
  points along the length to $\pm 0.05\,\text{mm}$. Record mass.
  Compute the actual fibre volume fraction $V_f$ from the known
  cloth areal mass, the number of plies, the measured laminate
  thickness, and the densities of fibre and matrix.
\end{itemize}

\textbf{Load test}.

\begin{itemize}[leftmargin=*]
\item Set up the three-point bending rig with the $250\,\text{mm}$
  support span. Load the beam at the centre, increasing the force
  gradually to failure. Record the peak force, the deflection at
  peak force, and the visible failure mode (tensile fracture on the
  bottom face, compressive crushing on the top face, delamination at
  the neutral axis, or a combination).
\item Compute the apparent flexural strength
  $\sigma_f = 3 F L / (2 b d^2)$ and the apparent flexural modulus
  $E_f = L^3 F / (4 b d^3 \delta)$ from the force-deflection record.
\end{itemize}

\textbf{Analysis}.

\begin{itemize}[leftmargin=*]
\item Compare the measured $E_f$ with the prediction from the rule of
  mixtures using known fibre and matrix moduli at the measured $V_f$.
  State the discrepancy and propose its sources.
\item Compare the measured $\sigma_f$ with the predicted longitudinal
  tensile strength from the rule of mixtures.
\item Photograph the fracture surface at $5\times$ to $20\times$
  magnification (a phone camera with a clip-on macro lens is
  sufficient). Identify whether the failure was matrix-dominated
  (matrix cracking, fibre pullout) or fibre-dominated (clean fibre
  fracture).
\end{itemize}

\textbf{Reflection ($1{,}500$ to $2{,}500$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
\item What fraction of the rule-of-mixtures prediction did the
  laminate achieve, and what is the dominant source of the gap?
\item What manufacturing variables most influenced the measured
  strength?
\item How would the design change for a beam intended to survive
  $10^7$ load cycles in service?
\item One feature of the failure mode that the reader had not
  anticipated.
\end{itemize}

\textbf{Deliverable}. Photographs of the lay-up and cure, the data
log, the analysis, the report.

\textbf{Time}. Six to twelve hours of lay-up and cure (most of which
is wait time during cure), four to six hours of testing and analysis,
four to six hours of write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A unidirectional carbon-fibre/epoxy laminate has $V_f = 0.6$, fibre
modulus $E_f = 230\,\text{GPa}$, matrix modulus $E_m = 3.5\,\text{GPa}$.
Compute the longitudinal modulus (Voigt average) and the transverse
modulus (Reuss bound). Estimate the transverse modulus using the
Halpin-Tsai equation with $\xi = 2$.
\end{exercise}

\begin{exercise}
A glass-epoxy laminate has fibre tensile strength $\sigma_f^* = 3.4\,
\text{GPa}$, matrix tensile strength $\sigma_m^* = 70\,\text{MPa}$,
fibre breaking strain $\varepsilon_f^* = 4.8\,\%$, matrix breaking
strain $\varepsilon_m^* = 5.0\,\%$, $V_f = 0.55$. Compute the
longitudinal tensile strength.
\end{exercise}

\begin{exercise}
A quasi-isotropic laminate of carbon-fibre/epoxy plies has $E_1 =
140\,\text{GPa}$ and $E_2 = 10\,\text{GPa}$ per ply. Compute the
quasi-isotropic modulus using the $3/8 (E_1 + E_2)$ approximation.
\end{exercise}

\begin{exercise}
A sandwich panel has face sheets of $1\,\text{mm}$ aluminium ($E_f =
70\,\text{GPa}$) separated by a $20\,\text{mm}$ aluminium-honeycomb
core. Compute the bending stiffness per unit width and compare with a
monolithic aluminium plate of the same total mass per unit area.
\end{exercise}

\begin{exercise}
A composite-overwrapped pressure vessel uses a cylindrical filament
winding at the optimum wind angle. Compute the optimum wind angle for
a closed-end cylindrical pressure vessel (where the hoop-to-axial
stress ratio is $2:1$) and explain why filament-wound tanks are not
fully optimised at this angle in practice.
\end{exercise}

\begin{exercise}
A carbon-fibre/epoxy $[0]_8$ laminate (eight unidirectional plies) is
loaded in tension along the fibre direction at $\sigma = 600\,\text{MPa}$.
A single ply is removed by a manufacturing defect. Compute the stress
in the remaining seven plies and compare with the laminate's static
strength.
\end{exercise}

\begin{exercise}
A glass-epoxy laminate is exposed to a humid environment and absorbs
$1.2\,\%$ moisture by mass at equilibrium. The matrix glass transition
temperature drops from $130\,\degC$ (dry) to $105\,\degC$ (wet).
Estimate the temperature derating for a service environment requiring
a $30\,\degC$ margin below $T_g$.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the Voigt (equal-strain) longitudinal modulus and the Reuss
(equal-stress) transverse modulus for a two-phase composite. State the
boundary conditions assumed in each case.
\end{exercise}

\begin{exercise}
Show that the bending stiffness of a sandwich panel of total thickness
$h$ and face-sheet thickness $t_f \ll h$ scales as $E_f t_f h^2 / 2$
in the thin-face limit, by neglecting the core's contribution to the
in-plane stiffness.
\end{exercise}

\begin{exercise}
Derive the optimum filament-wind angle for a closed-end cylindrical
pressure vessel by minimising fibre stress over the hoop and axial
directions. Show that the optimum is $\arctan(\sqrt{2}) \approx
54.7^\circ$.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the autoclave-pressure work required to consolidate a
$1\,\text{m}^2$ aerospace primary-structure panel from $V_f = 0.5$ to
$V_f = 0.6$. Compare with the autoclave's heating energy.
\end{exercise}

\begin{exercise}
Estimate the total fibre length in a commercial wind-turbine blade of
$60\,\text{m}$, using a working laminate of $40\,\text{mm}$ average
thickness, $V_f = 0.55$, and a fibre diameter of $15\,\si{\micro\meter}$.
\end{exercise}

\begin{exercise}
Estimate the BVID-induced reduction in compressive strength of a
quasi-isotropic carbon-epoxy laminate after a $5\,\text{J}$ impact,
using rule-of-thumb damage-tolerance allowables.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a classical-lamination-theory solver: given ply properties
$(E_1, E_2, G_{12}, \nu_{12}, t)$ and a lay-up sequence
$\{\theta_k, t_k\}$, compute the laminate's $\mathbf{A}$, $\mathbf{B}$,
$\mathbf{D}$ matrices and the in-plane and bending elastic constants.
Verify on a $[0/90]_s$ cross-ply and on a $[0/\pm 45/90]_s$ quasi-
isotropic laminate.
\end{exercise}

\begin{exercise}
Simulate the load-deflection response of a sandwich panel under
uniformly distributed load, using the Timoshenko shear-deformable beam
theory. Compute the deflection contributions from bending and from
shear and identify the span at which the two contributions are equal.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify a carbon-epoxy laminate lay-up for a primary-structure spar
cap that must carry $\pm 0.4\,\%$ longitudinal strain at design and
satisfy the ten-percent rule in each of four orientations. State the
ply counts and the lay-up sequence.
\end{exercise}

\begin{exercise}
Design a sandwich panel for a passenger-aircraft floor that supports
a uniform live load of $4\,\text{kPa}$ over a span of $0.6\,\text{m}$
with a deflection limit of $L/300$. Specify the face-sheet thickness,
the core thickness, and the core density.
\end{exercise}

\begin{exercise}
Specify a composite-overwrapped pressure vessel for breathing-air
storage at $30\,\text{MPa}$. State the cylinder material (aluminium
or thermoplastic), the overwrap fibre (carbon or glass), the wind
angle, and the safety factor against burst.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A composite-blade fragment from a failed wind-turbine blade is
recovered. The fracture surface shows clean fibre pullout on one side
and brittle matrix fracture on the other. Diagnose the failure mode,
the likely operating condition at failure, and the design or
manufacturing factor that limited the blade's life.
\end{exercise}

\begin{exercise}
A bicycle frame's carbon-epoxy seat tube exhibits a $20\,\text{mm}$
longitudinal crack after a moderate impact during transport. The
visible damage is barely audible by tap test. Estimate the residual
compressive strength of the tube and recommend whether the frame
should be retired or repaired.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A composite-overwrapped pressure vessel ruptures at $80\,\%$ of its
design pressure after eight months of static storage at full pressure.
Identify the dominant failure mechanism (stress rupture, environmental
attack, impact damage during installation) and propose a single
inspection or test that would distinguish them.
\end{exercise}

\begin{exercise}
A carbon-fibre fuselage panel develops a $50\,\text{mm}$ delamination
during ground operation, not in flight. Identify the likely cause and
the inspection-regime change that would prevent recurrence.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A vendor proposes a quasi-isotropic carbon-epoxy laminate that uses
$30\,\%$ of plies in the $0^\circ$ direction and the balance distributed
among $\pm 45^\circ$ and $90^\circ$. The structural-loading study shows
strong $0^\circ$ loading. Write a one-paragraph evaluation of the
proposal, identifying the constraint the proposed lay-up serves and
the trade-off it makes against pure $0^\circ$ loading.
\end{exercise}

\begin{exercise}
A start-up's pitch claims a recyclable thermoset matrix that
maintains aerospace-grade properties through five recycling cycles.
Identify the load-bearing properties most at risk after recycling and
the question to ask in the next investor meeting.
\end{exercise}
