\chapter{Trigonometry, complex numbers, phasors}
\label{vol02:ch03}

\epigraph{Complex numbers form a closed and self-consistent system,
in which the operations of addition, subtraction, multiplication,
and division can be carried out without exception.}{Gilbert Strang,
\textit{Introduction to Linear Algebra}, 5th edition,
Wellesley-Cambridge, 2016, paraphrased from the introduction to the
complex-number appendix \cite[appendix on complex
numbers]{text:strang2016}.}

\chapmeta{Half-life: foundational. The trigonometric identities,
Euler's identity, and the complex-number algebra are durable on the
century scale; the phasor convention and its use in steady-state AC
analysis are durable on the decade-to-century scale; the specific
instrumentation that produces phase measurements ages on the
decade scale. Archetypes: stability (rotation as preserved
structure, phasor magnitude as a conserved quantity over a complete
cycle). Exercise target: 35. Project track: analysis with optional
simulation, per Q55-vol02-ch03. Hazard class: none. Reader path
default: standard, with sections explicitly tagged. Named cases:
none required; the failure section discusses generic phase
wrap-around in working measurements.}

\noindent
A periodic signal carries two pieces of information at every
instant: a magnitude and a phase. The trigonometric apparatus the
reader brings from secondary school describes the magnitude through
the cosine and sine, and the phase through the angle in the cosine
argument. A second apparatus, the complex number, packages the same
two pieces into a single algebraic object whose multiplication rule
reproduces the addition of phase angles and the multiplication of
magnitudes simultaneously. The result is a working tool that turns
the differential equations of steady-state AC analysis into
ordinary algebra. We assemble the trigonometric identities the
reader will use across the next fifteen chapters, build the complex
numbers as ordered pairs of real numbers, state Euler's identity as
the bridge that connects the two pictures, and develop the phasor
as the engineer's working notation for sinusoidal signals at a
single frequency. The chapter's archetype is stability: rotation
preserves structure, and the phasor magnitude is a conserved
quantity over a complete cycle.

\section{Right triangles, the unit circle, identities}\pathtag{core}

The trigonometric functions in their working form arise from two
constructions that the reader has seen, possibly in different
notations, in secondary education. The first is the right triangle:
ratios of side lengths in a triangle with one $90^{\circ}$ angle.
The second is the unit circle: coordinates of a point that travels
around a circle of radius one. The two constructions agree on the
common range and disagree only in their natural domain. Engineering
uses both, sometimes in the same calculation.

\subsection{The ratios in a right triangle}

Consider a right triangle with the right angle at one vertex and an
acute angle $\theta$ at a second vertex. The side opposite $\theta$
has length $a$; the side adjacent to $\theta$ has length $b$; the
hypotenuse has length $c$. The Pythagorean theorem fixes
$a^{2} + b^{2} = c^{2}$. The six trigonometric ratios follow:
\begin{align*}
\sin\theta &= \frac{a}{c}, &
\cos\theta &= \frac{b}{c}, &
\tan\theta &= \frac{a}{b}, \\
\csc\theta &= \frac{c}{a}, &
\sec\theta &= \frac{c}{b}, &
\cot\theta &= \frac{b}{a}.
\end{align*}
The three on the left are the working set; the three on the right
are reciprocals included for completeness. An engineer fluent in
sine, cosine, and tangent rarely needs the reciprocals as separate
objects, although they appear in classical mechanics handbooks and
in the Taylor expansions to which Chapter 4 attends.

The right-triangle definitions have a natural domain of
$0 < \theta < \pi/2$. They do not extend to $\theta = 0$
or $\theta = \pi/2$ as triangle ratios because one of the legs
collapses. They do not extend to $\theta < 0$ or $\theta > \pi/2$
because a triangle has no obtuse interior angle alongside a right
angle. The unit circle removes both restrictions.

\subsection{The unit circle}

A unit circle in the Cartesian plane is the locus of points at
distance one from the origin: $\{(x, y) : x^{2} + y^{2} = 1\}$. We
parametrise the circle by the arc length $\theta$ traversed
counterclockwise from the point $(1, 0)$. The point on the circle
at parameter $\theta$ is
\[
(x, y) = (\cos\theta, \sin\theta).
\]
This single equation is the working definition of the cosine and
the sine for all real $\theta$. The right-triangle definitions
agree with the unit-circle definitions on $0 < \theta < \pi/2$, and
the unit-circle definitions extend to all angles through the
geometry of the circle. A point at parameter $\theta + 2\pi$ is the
same point as the one at $\theta$, so cosine and sine are
$2\pi$-periodic. A point at parameter $-\theta$ is the reflection
of the point at $\theta$ across the $x$-axis, so $\cos(-\theta) =
\cos\theta$ and $\sin(-\theta) = -\sin\theta$.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-unit-circle.tex}

The arc length is the radian. One full revolution is $2\pi$
radians, equal to $360^{\circ}$. The conversion $1 \,\text{rad} =
180^{\circ}/\pi \approx 57.296^{\circ}$ is worth memorising; the
inverse, $1^{\circ} = \pi/180 \,\text{rad} \approx 0.01745$, is
worth memorising too. Engineering practice uses radians inside
analytic formulas (where the small-angle approximations and the
Taylor series carry the natural unit) and degrees inside design
documents and field measurements (where the human reader does the
work). The two are interchangeable, but the conversion factor must
travel with the calculation.

\subsection{Symmetry and the canonical values}

The unit-circle definition makes several symmetry properties
visible. We collect them as identities the reader uses without
re-derivation. For all real $\theta$:
\begin{align*}
\cos(\theta + 2\pi) &= \cos\theta, &
\sin(\theta + 2\pi) &= \sin\theta, \\
\cos(-\theta) &= \cos\theta, &
\sin(-\theta) &= -\sin\theta, \\
\cos(\pi - \theta) &= -\cos\theta, &
\sin(\pi - \theta) &= \sin\theta, \\
\cos(\theta + \pi/2) &= -\sin\theta, &
\sin(\theta + \pi/2) &= \cos\theta.
\end{align*}
The last pair says that a quarter-cycle phase shift turns a cosine
into a (negated) sine and a sine into a cosine. The reader will
meet this fact in every section that treats two sinusoids in
quadrature.

The canonical values follow from the geometry of the equilateral
triangle (split to $30$-$60$-$90$) and the isoceles right triangle
($45$-$45$-$90$), together with the points $(1, 0)$, $(0, 1)$,
$(-1, 0)$, $(0, -1)$ on the axes. We tabulate the values most often
needed in working calculations:
\begin{center}
\begin{tabular}{cccccc}
\toprule
$\theta$ (rad) & $0$ & $\pi/6$ & $\pi/4$ & $\pi/3$ & $\pi/2$ \\
$\theta$ (deg) & $0$ & $30$ & $45$ & $60$ & $90$ \\
\midrule
$\sin\theta$ & $0$ & $1/2$ & $\sqrt{2}/2$ & $\sqrt{3}/2$ & $1$ \\
$\cos\theta$ & $1$ & $\sqrt{3}/2$ & $\sqrt{2}/2$ & $1/2$ & $0$ \\
$\tan\theta$ & $0$ & $\sqrt{3}/3$ & $1$ & $\sqrt{3}$ & undefined \\
\bottomrule
\end{tabular}
\end{center}
The reader is expected to reproduce the table from the geometry of
the unit circle without consulting a reference. The pattern $0$,
$1/2$, $\sqrt{2}/2$, $\sqrt{3}/2$, $1$ for the sine across the five
canonical angles is a memorable mnemonic that rewards five seconds
of inspection.

\paragraph{Worked example: reduction to the first quadrant.} The
canonical table covers $0$ to $\pi/2$. Every other angle reduces to
this range by the symmetry identities, and the reduction is the
working procedure for evaluating a sinusoid at any argument by hand.
Take $\cos(17\pi/6)$. First subtract whole revolutions: $17\pi/6 -
2\pi = 17\pi/6 - 12\pi/6 = 5\pi/6$, so $\cos(17\pi/6) = \cos(5\pi/6)$.
The angle $5\pi/6$ lies in the second quadrant, where cosine is
negative; the reference angle (the acute angle to the nearest part
of the real axis) is $\pi - 5\pi/6 = \pi/6$. Hence $\cos(5\pi/6) =
-\cos(\pi/6) = -\sqrt{3}/2$. The same three steps, reduce modulo
$2\pi$, identify the quadrant and its sign, reflect to the reference
angle, evaluate any trigonometric function at any argument without a
calculator. The procedure is worth drilling until it is automatic,
because the alternative (a calculator in degree mode when the
formula expects radians) is a recurring source of error in field
calculations.

\subsection{The Pythagorean and angle-sum identities}

Two families of identities carry most of the algebraic weight.

The \engterm{Pythagorean identity} comes from the unit circle
itself. Because $(\cos\theta)^{2} + (\sin\theta)^{2}$ is the square
of the distance from the origin to the point at parameter $\theta$,
and that distance is one,
\[
\cos^{2}\theta + \sin^{2}\theta = 1.
\]
Dividing by $\cos^{2}\theta$ gives $1 + \tan^{2}\theta =
\sec^{2}\theta$; dividing by $\sin^{2}\theta$ gives
$\cot^{2}\theta + 1 = \csc^{2}\theta$. The three forms each appear
in different working contexts.

The \engterm{angle-sum identities} are the algebraic content of
rotation. For any real angles $\alpha$ and $\beta$:
\begin{align*}
\cos(\alpha + \beta) &= \cos\alpha\cos\beta - \sin\alpha\sin\beta, \\
\sin(\alpha + \beta) &= \sin\alpha\cos\beta + \cos\alpha\sin\beta.
\end{align*}
A geometric proof draws two unit vectors at angles $\alpha$ and
$-\beta$, computes the cosine of the angle between them as a dot
product (the left-hand side of the first identity), and recognises
the right-hand side as the same dot product computed in
coordinates. We give the algebraic content here and the rotation
proof in section 3.5; the reader should be able to recover both.

The angle-difference identities are the angle-sum identities with
$\beta$ replaced by $-\beta$ and the symmetry of cosine and sine
applied:
\begin{align*}
\cos(\alpha - \beta) &= \cos\alpha\cos\beta + \sin\alpha\sin\beta, \\
\sin(\alpha - \beta) &= \sin\alpha\cos\beta - \cos\alpha\sin\beta.
\end{align*}
The double-angle identities are the angle-sum identities with
$\alpha = \beta = \theta$:
\begin{align*}
\cos(2\theta) &= \cos^{2}\theta - \sin^{2}\theta = 1 - 2\sin^{2}\theta = 2\cos^{2}\theta - 1, \\
\sin(2\theta) &= 2\sin\theta\cos\theta.
\end{align*}
The half-angle identities follow by solving the cosine
double-angle identity for $\sin^{2}\theta$ and $\cos^{2}\theta$:
\begin{align*}
\sin^{2}\theta &= \frac{1 - \cos(2\theta)}{2}, &
\cos^{2}\theta &= \frac{1 + \cos(2\theta)}{2}.
\end{align*}
The half-angle forms convert a squared sinusoid into a constant
plus a sinusoid at twice the frequency, and they are the working
identity behind the AC power calculation: the average of
$\cos^{2}(\omega t)$ over a full cycle is $1/2$, because the
$\cos(2\omega t)$ term integrates to zero.

\paragraph{Worked example: a power-average integral by half-angle.}
The half-angle identity does real work in the most common AC power
calculation. The instantaneous power into a resistor carrying
$i(t) = I_{m}\cos(\omega t)$ is $p(t) = R\,i^{2}(t) = R I_{m}^{2}
\cos^{2}(\omega t)$. Substitute the half-angle form $\cos^{2}(\omega
t) = \tfrac{1}{2}[1 + \cos(2\omega t)]$:
\[
p(t) = \tfrac{1}{2} R I_{m}^{2} + \tfrac{1}{2} R I_{m}^{2}\cos(2\omega t).
\]
The instantaneous power is a constant term plus a sinusoid at twice
the line frequency. Averaging over one period kills the
double-frequency term (a full cosine cycle integrates to zero),
leaving $P = \tfrac{1}{2} R I_{m}^{2}$. The factor of one half is the
average value of $\cos^{2}$, and it is the reason the root-mean-square
current $I_{\text{RMS}} = I_{m}/\sqrt{2}$ gives the heating power
through $P = I_{\text{RMS}}^{2} R$ with no extra factor. The
double-frequency ripple that the average discards is real: it is the
$100$- or $120$-hertz hum in a poorly filtered single-phase rectifier
and the torque pulsation that single-phase motors suffer and
three-phase motors do not.

\subsection{The small-angle approximations}

When the argument is small, the trigonometric functions collapse to
their leading Taylor terms. For $|\theta| \ll 1$ (radians, always
radians),
\[
\sin\theta \approx \theta, \qquad
\cos\theta \approx 1 - \tfrac{1}{2}\theta^{2}, \qquad
\tan\theta \approx \theta.
\]
The sine and tangent approximations are linear in the angle; the
cosine approximation keeps the quadratic term because its linear
term vanishes. These three approximations carry an enormous amount
of engineering weight. The pendulum's equation of motion is
nonlinear in the swing angle through a $\sin\theta$ restoring term;
replacing $\sin\theta$ by $\theta$ linearises it into the
simple-harmonic oscillator whose period is independent of amplitude.
The paraxial approximation of optics replaces $\sin\theta$ and
$\tan\theta$ by $\theta$ to make ray tracing through lenses linear.
The small-deflection theory of beams and plates replaces the exact
curvature by its linear approximation.

The cost of the approximation is bounded by the first dropped term.
The sine approximation $\sin\theta \approx \theta$ has relative error
$\theta^{2}/6$ to leading order. At $\theta = 0.1\,\text{rad}$
($\approx 5.7^{\circ}$) the relative error is about $1/600 \approx
0.17\%$; at $\theta = 0.2\,\text{rad}$ ($\approx 11.5^{\circ}$) it is
about $0.67\%$; at $\theta = 0.5\,\text{rad}$ ($\approx 28.6^{\circ}$)
it has grown to about $4.2\%$. The engineer who uses the small-angle
approximation states the angle range over which it holds and the
error it incurs, because the approximation degrades smoothly and a
silent extrapolation past its valid range is a recurring failure
(treated in section 3.7).

\subsection{Product-to-sum and sum-to-product}

The product of two sinusoids at different frequencies is a sum of
sinusoids at the sum and difference frequencies:
\begin{align*}
\cos\alpha\cos\beta &= \tfrac{1}{2}\left[\cos(\alpha - \beta) + \cos(\alpha + \beta)\right], \\
\sin\alpha\sin\beta &= \tfrac{1}{2}\left[\cos(\alpha - \beta) - \cos(\alpha + \beta)\right], \\
\sin\alpha\cos\beta &= \tfrac{1}{2}\left[\sin(\alpha + \beta) + \sin(\alpha - \beta)\right].
\end{align*}
These identities are the algebraic content of the
\engterm{heterodyne} principle: multiplying two sinusoids generates
content at the sum and difference of their frequencies, which is
the operating mechanism of the AM radio mixer, the lock-in
amplifier, the superheterodyne receiver, and the analogue
modulator. We use the identities here as algebra; the engineering
applications appear in Volume IV's wave chapters and in
Volume VII's communication chapters.

The reverse direction, sum to product, is also useful:
\begin{align*}
\cos\alpha + \cos\beta &= 2\cos\left(\tfrac{\alpha + \beta}{2}\right)\cos\left(\tfrac{\alpha - \beta}{2}\right), \\
\cos\alpha - \cos\beta &= -2\sin\left(\tfrac{\alpha + \beta}{2}\right)\sin\left(\tfrac{\alpha - \beta}{2}\right), \\
\sin\alpha + \sin\beta &= 2\sin\left(\tfrac{\alpha + \beta}{2}\right)\cos\left(\tfrac{\alpha - \beta}{2}\right).
\end{align*}
The first identity is the algebraic content of \engterm{beating}:
two sinusoids of nearly equal frequency $\alpha$ and $\beta$
combine into a single sinusoid at the average frequency
$(\alpha + \beta)/2$ whose amplitude is modulated at half the
difference frequency $(\alpha - \beta)/2$. The musician tuning two
strings hears the beat envelope; the engineer measuring the
difference between two oscillators uses the same envelope as a
high-resolution frequency comparison.
Figure~\ref{fig:vol02:ch03:beating} displays the construction for
a one-hertz frequency difference.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-beating.tex}

\begin{archetype}[Stability: rotation as preserved structure]
The unit circle is the canonical example of a stability archetype.
A point that moves around the circle preserves its distance from
the origin: the radius is a \engterm{conserved quantity} of the
motion. The angle changes, but the structure that the angle is an
angle on a circle of fixed radius does not. Engineering recognises
this pattern in many places: a phasor of constant magnitude and
varying phase, a rotor that turns without changing its diameter, an
energy-conserving oscillation that exchanges kinetic and potential
energy without dissipating either, a complex exponential
$e^{i\omega t}$ that traces the unit circle in the complex plane at
angular frequency $\omega$. The archetype recurs every time the
engineer needs a coordinate that distinguishes "where on the
cycle" from "how much."
\end{archetype}

\subsection{Inverse trigonometric functions}

The cosine, sine, and tangent are not invertible on their full
domains: each repeats every $2\pi$ (cosine and sine) or every
$\pi$ (tangent). To define an inverse, we restrict the domain to
a range on which the function is one-to-one and pick the inverse
that maps onto a chosen \engterm{principal range}. The standard
choices are
\begin{align*}
\arccos: [-1, 1] &\to [0, \pi], \\
\arcsin: [-1, 1] &\to [-\pi/2, \pi/2], \\
\arctan: \mathbb{R} &\to (-\pi/2, \pi/2).
\end{align*}
A computation that produces an angle through $\arctan$ recovers
only the principal value. Engineering practice often needs the
\engterm{full-range arctangent}, sometimes notated
$\operatorname{atan2}(y, x)$, which returns the angle of the point
$(x, y)$ in the range $(-\pi, \pi]$ by examining the signs of $x$
and $y$ separately. A complex-number magnitude-and-angle
calculation that uses $\arctan(y/x)$ without the sign check is
incorrect for points in the second and third quadrants. We return
to this issue in section 3.7 on phase wrap-around.

\section{Complex numbers as ordered pairs}\pathtag{core}

The complex numbers extend the real numbers by adjoining a single
new element, traditionally written $i$, that satisfies $i^{2} =
-1$. The construction can be motivated in several equivalent ways;
we use the ordered-pair construction because it makes the algebra
explicit and avoids any appeal to the mysterious "square root of
minus one."

\subsection{The set and the operations}

A \engterm{complex number} is an ordered pair $(a, b)$ of real
numbers. The set of complex numbers is $\mathbb{C} = \mathbb{R}
\times \mathbb{R}$. We define addition and multiplication on
$\mathbb{C}$ component-by-component for addition, and through a
specific rule for multiplication:
\begin{align*}
(a, b) + (c, d) &= (a + c, b + d), \\
(a, b) \cdot (c, d) &= (ac - bd, ad + bc).
\end{align*}
Addition is the same as the addition of two-dimensional vectors:
the sum is the diagonal of the parallelogram. Multiplication is
the construction whose meaning we develop below.

The pair $(a, 0)$ behaves under addition and multiplication exactly
like the real number $a$: $(a, 0) + (c, 0) = (a + c, 0)$ and
$(a, 0) \cdot (c, 0) = (ac, 0)$. We identify $(a, 0)$ with the
real number $a$ and treat the real numbers as a subset of the
complex numbers along the first axis.

The pair $(0, 1)$, multiplied by itself, gives
\[
(0, 1) \cdot (0, 1) = (0 \cdot 0 - 1 \cdot 1,\; 0 \cdot 1 + 1 \cdot 0) = (-1, 0).
\]
The square of $(0, 1)$ is the real number $-1$. We name $(0, 1)$
the \engterm{imaginary unit} and write it $i$:
\[
i = (0, 1), \qquad i^{2} = -1.
\]
The pair $(a, b)$ then equals $(a, 0) + (0, b) = a + b \cdot (0, 1)
= a + bi$. The traditional notation $a + bi$ is just the ordered
pair written with $1$ and $i$ as the basis vectors of a
two-dimensional real vector space. The number $a$ is the
\engterm{real part}, $\Re(z) = a$; the number $b$ is the
\engterm{imaginary part}, $\Im(z) = b$. The imaginary part of a
complex number is itself a real number; the name "imaginary" is
historical and refers to the symbol $i$, not to the value $b$.

The multiplication rule, restated in $a + bi$ notation, is
\[
(a + bi)(c + di) = ac + adi + bci + bdi^{2} = (ac - bd) + (ad + bc)i,
\]
which uses $i^{2} = -1$ and collects real and imaginary parts. The
ordered-pair definition and the $a + bi$ definition agree.

\subsection{The complex plane}

A complex number $z = a + bi$ is plotted as the point $(a, b)$ in
the Cartesian plane. The plane in this role is called the
\engterm{complex plane} or \engterm{Argand diagram}. The horizontal
axis is the \engterm{real axis}; the vertical axis is the
\engterm{imaginary axis}.

The \engterm{magnitude} (or modulus) of $z = a + bi$ is the
distance from the origin to the point $(a, b)$:
\[
|z| = \sqrt{a^{2} + b^{2}}.
\]
The \engterm{argument} (or phase) of $z$ is the angle from the
positive real axis to the line from the origin to $(a, b)$,
measured counterclockwise:
\[
\arg(z) = \theta, \qquad \tan\theta = \frac{b}{a},
\]
with the quadrant of $(a, b)$ disambiguating $\theta$ on
$(-\pi, \pi]$ via the full-range arctangent
$\operatorname{atan2}(b, a)$. The argument is undefined for $z = 0$
and is defined modulo $2\pi$ for every other $z$; engineering
practice picks a principal range, almost always $(-\pi, \pi]$ or
$[0, 2\pi)$.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-complex-plane.tex}

The \engterm{complex conjugate} of $z = a + bi$ is
\[
\bar{z} = a - bi.
\]
Geometrically, the conjugate is the reflection of $z$ across the
real axis (figure~\ref{fig:vol02:ch03:complex-plane}, panel (b)). The conjugate has the working property
\[
z \bar{z} = (a + bi)(a - bi) = a^{2} + b^{2} = |z|^{2}.
\]
This identity converts a complex magnitude squared into a real
arithmetic operation, which is the working tool for division.

\subsection{Division and rationalisation}

To divide one complex number by another,
\[
\frac{z}{w} = \frac{a + bi}{c + di},
\]
multiply numerator and denominator by the conjugate of the
denominator:
\[
\frac{a + bi}{c + di} = \frac{(a + bi)(c - di)}{(c + di)(c - di)} = \frac{(ac + bd) + (bc - ad)i}{c^{2} + d^{2}}.
\]
The denominator is now a positive real number; the numerator is a
complex number. The quotient is real-and-imaginary in the form
$x + yi$ that the engineer can read directly. This is the same
trick a secondary-school student learns to rationalise a real-valued
fraction with a square-root denominator; the conjugate plays the
role of the radical conjugate in the real case.

\paragraph{Worked example: division two ways.} Compute $(4 + 3i)/(2
- i)$ in rectangular form, then check the answer in polar form. By
rationalisation, multiply numerator and denominator by $2 + i$:
\[
\frac{4 + 3i}{2 - i} = \frac{(4 + 3i)(2 + i)}{(2 - i)(2 + i)}
= \frac{8 + 4i + 6i + 3i^{2}}{4 + 1}
= \frac{5 + 10i}{5} = 1 + 2i.
\]
The polar check converts each number first. The numerator has
magnitude $\sqrt{16 + 9} = 5$ and argument $\arctan(3/4) \approx
0.6435\,\text{rad}$; the denominator has magnitude $\sqrt{4 + 1} =
\sqrt{5}$ and argument $\arctan(-1/2) \approx -0.4636\,\text{rad}$.
Division subtracts the arguments and divides the magnitudes:
the quotient has magnitude $5/\sqrt{5} = \sqrt{5} \approx 2.236$ and
argument $0.6435 - (-0.4636) = 1.1071\,\text{rad}$. Converting back,
$\sqrt{5}(\cos 1.1071 + i\sin 1.1071) \approx \sqrt{5}(0.4472 +
0.8944 i) = 1 + 2i$, which matches. The two routes are the same
calculation seen from rectangular and polar coordinates; the
geometric reading (a quotient rotates by the argument difference and
scales by the magnitude ratio) is the one to carry forward
\cite[chapter 1]{text:v2ch03-needham-visual-complex}.

\subsection{Algebraic structure}

The complex numbers form a \engterm{field}: addition is
commutative and associative with identity $0$ and inverses given
by negation; multiplication is commutative and associative with
identity $1$ and inverses given by $1/z = \bar{z}/|z|^{2}$ for
$z \ne 0$; multiplication distributes over addition. Every
nonzero complex number has a multiplicative inverse, and the
inverse is given explicitly by the rationalisation construction
above. The reader does not need to verify the field axioms by
hand, but the working consequence is important: complex arithmetic
behaves under all the algebraic operations the reader expects from
real arithmetic, with one additional element ($i$) and one
additional rule ($i^{2} = -1$).

The complex numbers are also \engterm{algebraically closed}: every
non-constant polynomial with complex coefficients has a complex
root. The fundamental theorem of algebra (which we state without
proof; the proof belongs to a complex-analysis course) guarantees
that a polynomial of degree $n$ with complex coefficients has
exactly $n$ complex roots, counted with multiplicity. The real
numbers do not have this property: the polynomial $x^{2} + 1$ has
no real roots. The closure property is the deeper reason that
complex numbers appear in engineering: they are the smallest
algebraic structure in which the polynomials of linear-systems
theory always factor completely.

\section{Euler's identity, polar form}\pathtag{core}

The complex exponential $e^{i\theta}$, taken as a function of the
real variable $\theta$, traces the unit circle in the complex
plane. We state this as a fact and develop its working
consequences. The proof, which uses the Taylor series for
$e^{x}$, $\cos\theta$, and $\sin\theta$, belongs to Chapter 4 once
the series machinery is in place.

\subsection{The identity}

\textbf{Euler's identity.} For all real $\theta$,
\[
e^{i\theta} = \cos\theta + i\sin\theta.
\]
The left-hand side is the complex exponential evaluated at the
imaginary number $i\theta$. The right-hand side is a complex
number of magnitude one, plotted at angle $\theta$ on the unit
circle in the complex plane. The identity says that those two
expressions are the same complex number for every real $\theta$.

The reader is asked to take the identity as a working fact for
this chapter. The Taylor-series proof in Chapter 4 will show that
the series for $e^{x}$ at $x = i\theta$ separates into a real part
that matches the cosine series and an imaginary part that matches
the sine series; the identity then follows from term-by-term
comparison. The same proof shows that $e^{i\theta}$ inherits the
exponential's algebraic properties, in particular
\[
e^{i\alpha} \cdot e^{i\beta} = e^{i(\alpha + \beta)}.
\]
This identity is equivalent to the angle-sum identities for
cosine and sine: expand both sides using Euler's identity, equate
real and imaginary parts, and the angle-sum identities fall out.
We will use this fact below.

A famous specialisation. Setting $\theta = \pi$ gives
$e^{i\pi} = \cos\pi + i\sin\pi = -1 + 0 \cdot i = -1$, or
\[
e^{i\pi} + 1 = 0.
\]
This identity, sometimes called Euler's equation, packages five
fundamental constants ($0$, $1$, $i$, $\pi$, $e$) into a single
algebraic statement. It is a memorable consequence of the
underlying construction; engineering uses Euler's identity in the
general form $e^{i\theta} = \cos\theta + i\sin\theta$ rather than
the specialisation.

\subsection{Polar form}

Euler's identity gives every nonzero complex number a second
representation. A complex number $z = a + bi$ with magnitude
$|z| = r$ and argument $\arg(z) = \theta$ can be written
\[
z = a + bi = r\cos\theta + ir\sin\theta = r(\cos\theta + i\sin\theta) = r e^{i\theta}.
\]
The expression $z = r e^{i\theta}$ is the \engterm{polar form} of
$z$. The pair $(r, \theta)$ carries the same information as the
pair $(a, b)$, with the conversions
\begin{align*}
r &= \sqrt{a^{2} + b^{2}}, & \theta &= \operatorname{atan2}(b, a), \\
a &= r\cos\theta, & b &= r\sin\theta.
\end{align*}
The form $a + bi$ is the \engterm{rectangular form} or
\engterm{Cartesian form}; the form $r e^{i\theta}$ is the
\engterm{polar form} or \engterm{exponential form}.

The two forms are useful for different operations. Addition is
easy in rectangular form: $(a + bi) + (c + di) = (a + c) +
(b + d)i$. Multiplication is easy in polar form:
\[
(r_{1} e^{i\theta_{1}}) \cdot (r_{2} e^{i\theta_{2}}) = r_{1} r_{2} e^{i(\theta_{1} + \theta_{2})}.
\]
Multiplying complex numbers in polar form multiplies their
magnitudes and adds their phases. This is the algebraic content
of the rotation picture: a unit complex number $e^{i\theta}$
multiplies any complex number it meets by rotating it through
$\theta$ and leaving its magnitude unchanged. We return to this
in section 3.5.

Division in polar form is also a one-line operation:
\[
\frac{r_{1} e^{i\theta_{1}}}{r_{2} e^{i\theta_{2}}} = \frac{r_{1}}{r_{2}} e^{i(\theta_{1} - \theta_{2})}.
\]
The engineer doing complex arithmetic at scale converts to polar
form before multiplying or dividing and converts back to
rectangular form before adding or subtracting.

\subsection{Roots and powers}

Powers of a complex number in polar form are immediate. By
induction on the multiplication rule,
\[
z^{n} = (r e^{i\theta})^{n} = r^{n} e^{in\theta}
\]
for every integer $n$, with the convention $z^{0} = 1$ and
$z^{-1} = (1/r) e^{-i\theta}$. The identity $(\cos\theta +
i\sin\theta)^{n} = \cos(n\theta) + i\sin(n\theta)$ that follows
from this is \engterm{De Moivre's formula}; equating real and
imaginary parts gives the multiple-angle formulas for cosine and
sine without further work.

\paragraph{Worked example: the triple-angle formulas by De Moivre.}
Expand $(\cos\theta + i\sin\theta)^{3}$ two ways. By De Moivre, the
left-hand side equals $\cos(3\theta) + i\sin(3\theta)$. By the
binomial theorem, writing $c = \cos\theta$ and $s = \sin\theta$,
\[
(c + is)^{3} = c^{3} + 3c^{2}(is) + 3c(is)^{2} + (is)^{3}
= (c^{3} - 3cs^{2}) + i(3c^{2}s - s^{3}).
\]
Equating real parts gives $\cos(3\theta) = c^{3} - 3cs^{2} =
4\cos^{3}\theta - 3\cos\theta$ (using $s^{2} = 1 - c^{2}$). Equating
imaginary parts gives $\sin(3\theta) = 3c^{2}s - s^{3} = 3\sin\theta
- 4\sin^{3}\theta$. The triple-angle formulas, which are tedious to
derive by repeated application of the angle-sum identities, fall out
of one binomial expansion. The same method gives the $n$-fold
multiple-angle formula for any $n$ by expanding $(c + is)^{n}$ and
reading off the real and imaginary parts; this is the working route
to the Chebyshev polynomials, which are exactly $T_{n}(\cos\theta) =
\cos(n\theta)$.

Roots are slightly subtler because the argument is defined modulo
$2\pi$. The $n$ \engterm{$n$th roots} of a nonzero complex number
$z = r e^{i\theta}$ are the complex numbers
\[
w_{k} = r^{1/n} e^{i(\theta + 2\pi k)/n}, \qquad k = 0, 1, \ldots, n-1.
\]
The roots have a common magnitude $r^{1/n}$ and arguments spaced
by $2\pi/n$ around the circle of radius $r^{1/n}$. The $n$ roots
of unity, that is, the $n$th roots of $z = 1$, sit at the corners
of a regular $n$-gon inscribed in the unit circle. The engineer
who needs to design a balanced $n$-phase electrical system or to
compute a discrete Fourier transform of length $n$ is using the
$n$th roots of unity as the working basis.
Figure~\ref{fig:vol02:ch03:roots-unity} displays the construction
for $n = 6$ and highlights the three cube roots of unity as the
phasors of a balanced three-phase system.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-roots-unity.tex}

\paragraph{Worked example: the three cube roots of $8$.} Find all
complex $w$ with $w^{3} = 8$. Write the right-hand side in polar
form as $8 = 8\,e^{i\cdot 0}$, recognising that the argument is
defined modulo $2\pi$ so that $8 = 8\,e^{i 2\pi k}$ for every
integer $k$. The cube roots have magnitude $8^{1/3} = 2$ and
arguments $(0 + 2\pi k)/3$ for $k = 0, 1, 2$:
\[
w_{0} = 2\,e^{i 0} = 2, \quad
w_{1} = 2\,e^{i 2\pi/3} = -1 + i\sqrt{3}, \quad
w_{2} = 2\,e^{i 4\pi/3} = -1 - i\sqrt{3}.
\]
A reader who stops at $w_{0} = 2$, the only real cube root, has
found one third of the answer. The two complex roots are the ones
the fundamental theorem of algebra guarantees and the ones a
real-number-only search misses. The three roots sum to zero (the
sum of the cube roots of any nonzero number is zero) and sit at the
corners of an equilateral triangle inscribed in the circle of
radius $2$. The general construction, and the geometric reason the
roots tile the circle evenly, is developed in
\cite[chapter 1]{text:v2ch03-stein-shakarchi-complex}.

\paragraph{Worked example: roots of a complex polynomial.} The
$n$th-root construction handles a polynomial of the special form
$z^{n} = w$. A general polynomial needs the standard factoring tools,
and the complex plane is the setting in which they always close. Find
all roots of
\[
P(z) = z^{4} - 2z^{2} + 4 = 0.
\]
The polynomial is quadratic in $u = z^{2}$: $u^{2} - 2u + 4 = 0$, so
$u = 1 \pm \sqrt{1 - 4} = 1 \pm \sqrt{-3} = 1 \pm i\sqrt{3}$. The two
values of $u$ are complex conjugates with magnitude $|u| = \sqrt{1 + 3}
= 2$ and arguments $\pm\pi/3$ (because $1 + i\sqrt 3 = 2e^{i\pi/3}$).
Each value of $u$ contributes two square roots $z = \pm u^{1/2}$. Taking
$u = 2e^{i\pi/3}$, the square roots are $\sqrt{2}\,e^{i\pi/6}$ and
$\sqrt{2}\,e^{i(\pi/6 + \pi)} = \sqrt{2}\,e^{i 7\pi/6}$; taking the
conjugate $u = 2e^{-i\pi/3}$, the square roots are
$\sqrt{2}\,e^{-i\pi/6}$ and $\sqrt{2}\,e^{-i 7\pi/6}$. The four roots
are
\[
z = \sqrt{2}\,e^{\pm i\pi/6}, \qquad z = \sqrt{2}\,e^{\pm i 5\pi/6},
\]
where the second pair rewrites $7\pi/6$ and $-7\pi/6$ in the principal
range as $-5\pi/6$ and $5\pi/6$. The four roots have common magnitude
$\sqrt{2}$ and sit symmetrically about both axes, as they must: a
polynomial with real coefficients has roots that occur in conjugate
pairs, so the root set is symmetric across the real axis. The reader
can reconstruct $P(z)$ as the product of two real quadratic factors,
$(z^{2} - \sqrt{2}\sqrt{3}\,z + 2)(z^{2} + \sqrt{2}\sqrt{3}\,z + 2)$,
each carrying one conjugate pair, which is the factorisation a circuit
or control engineer reads as a pair of second-order subsystems. The
roots are the poles of a fourth-order transfer function; their
common magnitude is the undamped natural frequency and their argument
off the imaginary axis is the damping, the reading developed in
Volume V \cite[chapter 1]{text:v2ch03-stein-shakarchi-complex}.

\subsection{The exponential as a working tool}

The complex exponential turns the algebra of sinusoids into the
algebra of exponentials, and the algebra of exponentials is one
of the cleanest algebras in mathematics. We collect the working
consequences for use in the rest of the chapter.

\begin{itemize}[leftmargin=*]
  \item Real and imaginary parts of $e^{i\theta}$:
    $\Re(e^{i\theta}) = \cos\theta$, $\Im(e^{i\theta}) = \sin\theta$.
  \item Magnitude: $|e^{i\theta}| = 1$ for all real $\theta$.
  \item Conjugate: $\overline{e^{i\theta}} = e^{-i\theta}$.
  \item Symmetric formulas (the \engterm{Euler formulas} for
    cosine and sine):
    \[
    \cos\theta = \tfrac{1}{2}\left(e^{i\theta} + e^{-i\theta}\right),
    \qquad
    \sin\theta = \tfrac{1}{2i}\left(e^{i\theta} - e^{-i\theta}\right).
    \]
  \item Time evolution at angular frequency $\omega$:
    $e^{i\omega t}$ traces the unit circle counterclockwise once
    per period $T = 2\pi/\omega$.
\end{itemize}
The Euler formulas express cosine and sine as combinations of
complex exponentials. The reader who memorises them recovers any
trigonometric identity by manipulating exponentials and reading
off real and imaginary parts. The angle-sum identities, the
product-to-sum identities, and the multiple-angle identities all
fall out of $e^{i\alpha} e^{i\beta} = e^{i(\alpha + \beta)}$ and
its consequences.

\begin{estimation}
\textbf{Question.} A complex number is given as $z = 3 + 4i$.
Estimate, before computing, the magnitude and the argument of
$z$ to within ten percent and to within ten degrees.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The magnitude is the distance from the origin to the point
$(3, 4)$. The Pythagorean triangle $3$-$4$-$5$ is the
secondary-school stand-by; the magnitude is exactly $5$, and an
estimator who notices the triangle pays no time at all. An
estimator who does not notice the triangle observes that
$|z|^{2} = 9 + 16 = 25$ and reports $|z| = 5$.

The argument is the angle from the positive real axis to the
point $(3, 4)$ in the first quadrant. The point lies above the
$y = x$ line (because $4 > 3$), so the angle is between $\pi/4$
($45^{\circ}$) and $\pi/2$ ($90^{\circ}$). The slope $4/3 \approx
1.33$ corresponds to an angle whose tangent is $1.33$, between
$45^{\circ}$ ($\tan = 1$) and $60^{\circ}$ ($\tan = \sqrt{3}
\approx 1.73$); a linear interpolation puts the estimate near
$53^{\circ}$ or $0.93 \,\text{rad}$. The actual value is
$\arctan(4/3) = 0.9273 \,\text{rad} = 53.13^{\circ}$, within a
fraction of a degree of the estimate.

The lesson is procedural. A complex number's magnitude is the
hypotenuse of a right triangle, and a complex number's argument
is the angle of that triangle. Estimating both before computing
keeps a sign error or a quadrant error from passing unnoticed.
\end{estimation}

\subsection{Why engineering uses the complex exponential}

The reader who arrives at this section from a pure-mathematics
sequence may wonder why engineering bothers with the complex
exponential when the cosine and sine carry all the information.
The answer is operational. Linear time-invariant systems (the
class to which AC steady-state circuits, mass-spring-damper
mechanical systems, RC and RL networks, and small-signal
amplifiers all belong) respond to a complex-exponential input
$e^{i\omega t}$ by producing a complex-exponential output
$H(\omega) e^{i\omega t}$, where $H(\omega)$ is a complex number
that depends on the system and the frequency. The factor
$H(\omega)$ is the \engterm{transfer function} at frequency
$\omega$. Its magnitude is the system's amplitude gain at that
frequency; its argument is the system's phase shift at that
frequency.

Doing the same calculation in cosine-sine notation produces a
pair of coupled equations for the in-phase and the quadrature
components, with the angle-sum identities applied at every
multiplication. The complex-exponential calculation produces a
single equation in a single complex variable, with the
multiplications carried out as ordinary algebra. The two
calculations agree because they are the same calculation; the
complex exponential is the bookkeeping that keeps the in-phase
and the quadrature components together. We make this concrete in
the next section.

\begin{mastery}
\textbf{Analytic functions and the Cauchy-Riemann equations.} A
complex function $f(z)$ is \engterm{differentiable} at $z_{0}$ if the
limit $f'(z_{0}) = \lim_{h \to 0}[f(z_{0} + h) - f(z_{0})]/h$ exists
and takes the same value no matter how the complex increment $h$
approaches zero. The two-dimensional freedom in the approach direction
makes complex differentiability a far stronger condition than its
real one-dimensional analogue. Write $f(x + iy) = u(x, y) + i v(x, y)$
with real $u$ and $v$. Approaching along the real axis ($h$ real) and
along the imaginary axis ($h$ imaginary) and demanding the two limits
agree gives the \engterm{Cauchy-Riemann equations}
\[
\frac{\partial u}{\partial x} = \frac{\partial v}{\partial y}, \qquad
\frac{\partial u}{\partial y} = -\frac{\partial v}{\partial x}.
\]
A function that satisfies them on an open region (with continuous
partials) is \engterm{analytic} there. Two consequences carry
engineering weight. First, $u$ and $v$ each satisfy Laplace's equation
$\nabla^{2} u = 0$, so the real and imaginary parts of an analytic
function are harmonic; this is the route by which complex variables
solve two-dimensional electrostatics, ideal fluid flow, and
steady-state heat conduction. Second, an analytic function with
$f'(z_{0}) \ne 0$ preserves angles, the conformal property of the next
box. The exponential $e^{z}$, the polynomials, and the rational
functions of linear-systems theory are analytic wherever their
denominators are nonzero; the points where a rational transfer
function fails to be analytic are its poles, and the residue
calculus that evaluates contour integrals around those poles is the
working machinery of the inverse Laplace transform
\cite[chapter 2]{text:v2ch03-stein-shakarchi-complex}.
\end{mastery}

\begin{mastery}
\textbf{Conformal maps and the Mobius transformation.} An analytic
function with nonzero derivative maps a small figure near $z_{0}$ to a
similar figure near $f(z_{0})$: it rotates by $\arg f'(z_{0})$ and
scales by $|f'(z_{0})|$, and it preserves the angle between any two
curves that cross. A map with this angle-preserving property is
\engterm{conformal}. The square map $w = z^{2}$ carries the orthogonal
Cartesian grid to two families of mutually orthogonal parabolas, drawn
in figure~\ref{fig:vol02:ch03:conformal-map}; the right angles of the
grid survive as right angles between the image curves. The most useful
conformal map in engineering is the \engterm{Mobius transformation}
$w = (az + b)/(cz + d)$ with $ad - bc \ne 0$, which maps circles and
lines to circles and lines and is the geometry behind the Smith chart
of radio-frequency impedance matching: the Smith chart is the image of
the impedance half-plane under the Mobius map
$\Gamma = (Z - Z_{0})/(Z + Z_{0})$ that sends impedance to reflection
coefficient. The conformal map turns a hard boundary-value problem on
an awkward region into an easy one on a disk or a half-plane, solves it
there, and maps the solution back; the technique is the working tool
of two-dimensional potential theory
\cite[chapter 6]{text:v2ch03-needham-visual-complex}.
\end{mastery}

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-conformal-map.tex}

\section{Phasors as the engineer's complex-number tool}\pathtag{core}

The phasor is the working notation that engineering uses to
represent a sinusoidal signal at a single known frequency by a
single complex number. We define the phasor, state its working
relationship to the time-domain signal, and demonstrate the
operations of phasor algebra on the standard linear circuit
elements.

\subsection{Definition}

Consider a sinusoidal signal of amplitude $A$, angular frequency
$\omega$, and initial phase $\varphi$:
\[
x(t) = A \cos(\omega t + \varphi).
\]
The \engterm{phasor} representing $x(t)$ is the complex number
\[
\tilde{X} = A e^{i\varphi}.
\]
The phasor carries the amplitude $A$ as its magnitude and the
initial phase $\varphi$ as its argument. It does not carry the
frequency $\omega$. The frequency is shared by every signal in a
single-frequency steady-state analysis, and dropping it from the
notation removes a redundancy without losing information.

The relationship between the time-domain signal and the phasor is
\[
x(t) = \Re\left[\tilde{X} e^{i\omega t}\right],
\]
which the reader should verify by expanding:
\[
\Re\left[A e^{i\varphi} e^{i\omega t}\right] = \Re\left[A e^{i(\omega t + \varphi)}\right] = A\cos(\omega t + \varphi).
\]
The first equality uses the exponential addition rule; the second
uses the real part of Euler's identity. The phasor and the
time-domain signal are interchangeable representations of the
same content at the same frequency.

A note on conventions. Some textbooks use the sine convention
$x(t) = A \sin(\omega t + \varphi)$ and define the phasor as the
same $A e^{i\varphi}$, with the time-domain recovery rule using
the imaginary part instead of the real part. The two conventions
agree on amplitude and phase up to a $\pi/2$ shift in the
reference signal; what matters is internal consistency within a
calculation. This book uses the cosine convention.

A second note. Some textbooks define the phasor with an
amplitude that is the RMS value $A/\sqrt{2}$ of the sinusoid
rather than the peak value $A$. The convention matters for power
calculations, where the RMS phasor gives the average power
through $P = \Re[\tilde{V}\bar{\tilde{I}}]$ directly without an
extra factor of one half. We use the peak-value convention in
this chapter and note where the RMS convention would simplify a
power expression.

\subsection{Linearity and superposition}

The phasor representation is linear: a sum of two sinusoids at
the same frequency is represented by the sum of their phasors.
Concretely, if $x_{1}(t) = A_{1}\cos(\omega t + \varphi_{1})$ and
$x_{2}(t) = A_{2}\cos(\omega t + \varphi_{2})$, then
\[
x_{1}(t) + x_{2}(t) = \Re\left[(\tilde{X}_{1} + \tilde{X}_{2}) e^{i\omega t}\right],
\]
so the phasor of the sum is $\tilde{X}_{1} + \tilde{X}_{2}$. The
sum is computed in the complex plane by ordinary vector addition.

The result of adding two sinusoids at the same frequency is
another sinusoid at the same frequency, with amplitude and phase
read off the sum phasor's magnitude and argument. The reader who
has tried to add two cosines with different phases by trigonometric
identity has seen how messy that calculation is; the phasor
calculation is one line.

A worked example: add $x_{1}(t) = 3\cos(\omega t)$ and
$x_{2}(t) = 4\cos(\omega t + \pi/2)$. The phasors are
$\tilde{X}_{1} = 3$ and $\tilde{X}_{2} = 4 e^{i\pi/2} = 4i$. The
sum is $\tilde{X}_{1} + \tilde{X}_{2} = 3 + 4i$, with magnitude
$|3 + 4i| = 5$ and argument $\arctan(4/3) \approx 0.9273
\,\text{rad} \approx 53.13^{\circ}$. The summed signal is therefore
$x_{1}(t) + x_{2}(t) = 5\cos(\omega t + 0.9273)$. The reader who
verifies this by expanding the right-hand side and comparing to
the left-hand side will recover the same answer through several
applications of the angle-sum identities.

\subsection{Differentiation in the phasor domain}

The derivative of a sinusoid is a sinusoid at the same frequency,
phase-shifted by $\pi/2$. In phasor notation,
\[
\frac{d}{dt}\Re[\tilde{X} e^{i\omega t}] = \Re\left[i\omega \tilde{X} e^{i\omega t}\right].
\]
The derivative operator in the time domain corresponds to
multiplication by $i\omega$ in the phasor domain. The integral
operator corresponds to division by $i\omega$. This is the central
operational fact of phasor analysis: differential equations of
linear time-invariant systems become algebraic equations in the
phasor variable. We accept the differentiation rule here on the
strength of the trigonometric identity
$d/dt[\cos(\omega t + \varphi)] = -\omega\sin(\omega t + \varphi) =
\omega\cos(\omega t + \varphi + \pi/2)$, which the reader can
verify by inspection. The formal calculus arrives in Chapter 5;
the phasor result holds independently of how the calculus is
formalised, because the same identity expresses it.

\subsection{Impedance}

Phasor analysis applied to a resistor, an inductor, and a
capacitor produces three working impedances. The resistor obeys
Ohm's law in the time domain: $v_{R}(t) = R i_{R}(t)$. In phasor
notation, $\tilde{V}_{R} = R \tilde{I}_{R}$, so the resistor's
\engterm{impedance} is the real number
\[
Z_{R} = R.
\]
The voltage and current are in phase across a resistor; their
phasors point in the same direction in the complex plane.

The inductor obeys $v_{L}(t) = L \, di_{L}(t)/dt$. In phasor
notation, $\tilde{V}_{L} = L \cdot i\omega \tilde{I}_{L} = i\omega
L \tilde{I}_{L}$, so the inductor's impedance is
\[
Z_{L} = i\omega L.
\]
The voltage leads the current by $\pi/2$ ($90^{\circ}$) across an
inductor; the voltage phasor is the current phasor rotated
counterclockwise by a quarter turn.

The capacitor obeys $i_{C}(t) = C \, dv_{C}(t)/dt$. In phasor
notation, $\tilde{I}_{C} = i\omega C \tilde{V}_{C}$, so the
capacitor's impedance is
\[
Z_{C} = \frac{1}{i\omega C} = -\frac{i}{\omega C}.
\]
The current leads the voltage by $\pi/2$ across a capacitor; the
current phasor is the voltage phasor rotated counterclockwise by
a quarter turn (which, equivalently, says that the voltage phasor
is the current phasor rotated clockwise by a quarter turn).

The three impedances together let the engineer treat any
combination of resistors, inductors, and capacitors as a single
network of complex impedances and apply Kirchhoff's laws as
phasor algebra. Series impedances add: $Z_{\text{series}} =
Z_{1} + Z_{2}$. Parallel impedances combine through the
reciprocal sum: $1/Z_{\text{parallel}} = 1/Z_{1} + 1/Z_{2}$. Ohm's
law in phasor form, $\tilde{V} = Z \tilde{I}$, applies to any
two-terminal element or network.
The real part of an impedance is the \engterm{resistance} and the
imaginary part is the \engterm{reactance}; the impedance of a series
RLC branch is $Z = R + iX$ with $X = \omega L - 1/(\omega C)$, drawn
as the impedance triangle of
figure~\ref{fig:vol02:ch03:impedance-triangle}. A positive net
reactance (inductive) puts the impedance in the first quadrant and
makes the current lag; a negative net reactance (capacitive) puts it
in the fourth quadrant and makes the current lead.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-impedance-triangle.tex}

\paragraph{Worked example: a parallel RC branch.} A resistor $R =
1\,\si{\kilo\ohm}$ in parallel with a capacitor $C = 0.1\,\si
{\micro\farad}$ is driven at $\omega = 10^{4}\,\si{\radian\per
\second}$. The branch impedance combines through the reciprocal sum.
The capacitor impedance is $Z_{C} = 1/(i\omega C) = 1/(i \cdot 10^{4}
\cdot 10^{-7}) = 1/(i \cdot 10^{-3}) = -1000 i\,\si{\ohm}$, a
capacitive reactance of $1\,\si{\kilo\ohm}$. The parallel
combination is
\[
Z = \frac{R \cdot Z_{C}}{R + Z_{C}}
= \frac{1000 \cdot (-1000 i)}{1000 - 1000 i}
= \frac{-10^{6} i}{1000(1 - i)}
= \frac{-1000 i}{1 - i}.
\]
Rationalise by $1 + i$: $Z = -1000 i (1 + i)/((1 - i)(1 + i)) =
-1000 i(1 + i)/2 = -1000(i + i^{2})/2 = -1000(i - 1)/2 = 500 - 500 i
\,\si{\ohm}$. The branch looks like a $500\,\si{\ohm}$ resistance in
series with a $500\,\si{\ohm}$ capacitive reactance at this
frequency, with magnitude $|Z| = 500\sqrt{2} \approx 707\,\si{\ohm}$
and phase $-\pi/4$. The reader who tries the same calculation in the
time domain confronts a first-order differential equation; the
phasor calculation is four lines of complex arithmetic. The
parallel-impedance reciprocal sum and the resulting frequency
dependence are the working content of the first-order RC filter,
treated at length in \cite[chapter 14]{text:v2ch03-nilsson-riedel}.

\subsection{Power in the phasor representation}

Average power dissipated in a resistor over a complete AC cycle is
the time average of $i^{2}(t) R$. For a sinusoidal current
$i(t) = I_{m}\cos(\omega t + \varphi)$, the time average of
$\cos^{2}$ over a cycle is $1/2$ (from the half-angle identity),
so the average power is $P = I_{m}^{2} R / 2$. In phasor
notation with the peak-value convention,
\[
P = \tfrac{1}{2}\Re\left[\tilde{V}\bar{\tilde{I}}\right].
\]
The factor of one half is the cycle-average of $\cos^{2}$. With
the RMS-value convention for phasors ($\tilde{V}$ scaled by
$1/\sqrt{2}$), the formula becomes $P = \Re[\tilde{V}\bar{\tilde{I}}]$
with no factor.

The full \engterm{complex power} $S = \tfrac{1}{2}\tilde{V}\bar{\tilde{I}}$
has real and imaginary parts. The real part is the
\engterm{average power} $P$ (in watts), the rate at which
energy is dissipated in resistive elements. The imaginary part is
the \engterm{reactive power} $Q$ (in volt-amperes reactive, var),
the rate at which energy is exchanged between the source and the
inductive or capacitive elements without net dissipation. The
magnitude $|S|$ is the \engterm{apparent power} (in volt-amperes,
VA). The ratio $P/|S| = \cos\varphi$, where $\varphi$ is the
phase angle between voltage and current, is the
\engterm{power factor}. A purely resistive load has power factor
one; a purely reactive load has power factor zero. Industrial
power systems care about the power factor because the apparent
power, not the average power, sets the conductor and transformer
sizing. Figure~\ref{fig:vol02:ch03:power-triangle} draws the three
powers as the sides of the \engterm{power triangle} and shows what
power-factor correction does to it.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-power-triangle.tex}

\begin{estimation}
\textbf{Question.} A small industrial workshop draws an apparent
power of $S = 30 \,\si{\kilo\text{VA}}$ at $400\,\si{\volt}$
line-to-line, $50\,\si{\hertz}$, at a power factor of
$\cos\varphi = 0.78$ lagging (typical of a workshop running
induction motors without correction, current as of 2026). Estimate,
before computing, the average power $P$ in kilowatts, the
reactive power $Q$ in kilovolt-amperes reactive, and the
capacitance $C$ in microfarads that would have to be connected in
parallel to bring the power factor to $0.95$ lagging.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The average power is $P = S\cos\varphi$, so $P \approx 30 \cdot
0.78 = 23.4 \,\si{\kilo\watt}$. Twenty-three kilowatts is the
real load on the supply, and the supply transformer and the
conductors must be sized for the apparent thirty kilovolt-amperes
because that is the magnitude of the current the supply actually
delivers. The reactive power is $Q = S\sin\varphi$ with
$\sin\varphi = \sqrt{1 - 0.78^{2}} \approx 0.626$, so $Q \approx
30 \cdot 0.626 = 18.8 \,\si{\kilo\text{var}}$.

To raise the power factor to $\cos\varphi' = 0.95$ at the same
real power $P$, the new reactive power is $Q' = P\tan\varphi'
= P \sin\varphi'/\cos\varphi'$. We have $\sin\varphi' = \sqrt{1 -
0.95^{2}} \approx 0.312$, so $\tan\varphi' \approx 0.329$ and
$Q' \approx 23.4 \cdot 0.329 \approx 7.7 \,\si{\kilo\text{var}}$.
The capacitor must absorb the difference $\Delta Q = Q - Q'
\approx 18.8 - 7.7 = 11.1 \,\si{\kilo\text{var}}$ of leading
reactive power. A three-phase capacitor bank delivering
$11.1 \,\si{\kilo\text{var}}$ at $400\,\si{\volt}$ line-to-line
(equivalently $400/\sqrt{3} \approx 231\,\si{\volt}$
line-to-neutral) with $\omega = 2\pi \cdot 50 \approx 314 \,\si
{\radian\per\second}$ requires $C = \Delta Q / (\omega V_{LL}^{2})
\approx 11{,}100 / (314 \cdot 400^{2}) \approx 2.2 \times 10^{-4}
\,\si{\farad} = 220 \,\si{\micro\farad}$ total across the three
delta-connected legs. An estimator who reaches for the
correction-capacitor formula directly arrives at the same value
in three lines.

The lesson is procedural. Power-factor correction is a phasor
calculation at a single frequency. The wattmeter on the supply
reports $P$; the apparent-power meter reports $|S|$; the
reactive-power meter reports $Q$. The capacitor's job is to
supply leading reactive power that cancels part of the load's
lagging reactive power, leaving the supply to deliver only the
residual. The transformer's rating shifts from the
uncorrected $30 \,\si{\kilo\text{VA}}$ to the corrected
$P/\cos\varphi' \approx 24.6 \,\si{\kilo\text{VA}}$, a reduction
of about $18\%$, which is why utilities charge a tariff penalty
for low power factor.
\end{estimation}

\subsection{What the phasor buys}

The phasor representation reduces the steady-state analysis of a
linear AC circuit to four steps. First, replace each element by
its impedance. Second, write Kirchhoff's voltage and current laws
in phasor form. Third, solve the resulting algebraic system for
the unknown phasors. Fourth, convert each phasor back to a
time-domain sinusoid. The differential equations of the original
circuit have been replaced by complex-valued linear algebra. The
project at the end of this chapter walks the reader through the
four steps for a series RLC circuit; we treat that example in
detail in section 3.8.

The price of the simplification is that the phasor representation
is exact only at a single frequency and only in steady state. A
transient response (the response during the first few cycles
after a switch closes) is not a single-frequency phenomenon and
needs the full differential equation. A signal with components at
multiple frequencies is a sum of phasors at each frequency, each
analysed separately and recombined; the linear-superposition
principle for LTI systems is what licences the recombination. A
nonlinear element (a diode, a saturating inductor, a clipping
amplifier) violates the assumptions; phasor analysis does not
apply.

\begin{mastery}
\textbf{The Laplace generalisation.} The phasor representation
$x(t) = \Re[\tilde X e^{i\omega t}]$ uses $i\omega$ as the
frequency variable. Replacing $i\omega$ with a general complex
$s = \sigma + i\omega$, and the steady-state input $e^{i\omega t}$
with the more general $e^{s t}$, yields the \engterm{Laplace
transform} of the differential equation. The transformed impedances
are $Z_{R}(s) = R$, $Z_{L}(s) = s L$, and $Z_{C}(s) = 1/(s C)$; the
algebraic technique of phasor analysis carries over with $s$ in
place of $i\omega$. The Laplace picture captures both the
steady-state response (set $s = i\omega$) and the transient
response (poles of the transfer function at $s = \sigma + i\omega$
with $\sigma < 0$ correspond to damped oscillations). The
machinery arrives in Volume V on dynamics and control; the phasor
representation of this chapter is its frequency-only specialisation
\cite{text:v2ch03-nilsson-riedel}.

\textbf{Negative frequencies and complex baseband.} The complex
exponential $e^{i\omega t}$ at $\omega < 0$ traces the unit circle
clockwise rather than counterclockwise. The cosine, written through
Euler's formula as $\cos(\omega t) = \tfrac{1}{2}(e^{i\omega t} +
e^{-i\omega t})$, contains equal energy at $+\omega$ and
$-\omega$. The two-sided spectrum of a real signal is
conjugate-symmetric around zero frequency; a plot of the
positive half discards a copy of the same information. Communication
systems that use quadrature modulation (in-phase and quadrature
components carried on $\cos(\omega t)$ and $\sin(\omega t)$) work
in the complex-baseband representation $z(t) = I(t) + i Q(t)$, in
which the negative frequencies carry independent information about
the modulating signal. Volume VII develops this picture.
\end{mastery}

\begin{mastery}
\textbf{The analytic signal and instantaneous phase.} The phasor
$\tilde X = A e^{i\varphi}$ assumes a single fixed frequency, a
single fixed amplitude, and a single fixed phase. A real signal whose
amplitude and phase drift slowly compared with the carrier period is
described instead by the \engterm{analytic signal}, the
complex-valued function whose real part is the measured signal and
whose imaginary part is the measured signal phase-shifted by
$-\pi/2$ at every frequency (the Hilbert transform). For a narrowband
signal $x(t) = A(t)\cos(\omega t + \varphi(t))$ with slowly-varying
$A(t)$ and $\varphi(t)$, the analytic signal is $z(t) = A(t)
e^{i(\omega t + \varphi(t))}$, and the \engterm{instantaneous
amplitude} $|z(t)| = A(t)$ and \engterm{instantaneous phase}
$\arg z(t) = \omega t + \varphi(t)$ recover the envelope and the
phase trajectory directly. The instantaneous frequency is the
derivative of the instantaneous phase. Vibration diagnostics,
communications demodulation, and seismic envelope analysis all run on
the analytic signal; it is the phasor generalised from a single
constant to a slowly-varying trajectory, and the wrap discipline of
section 3.7 applies to its phase exactly as it applies to the
constant phasor's argument.
\end{mastery}

\subsection{The Fourier series as a sum of phasors}

A single phasor represents a signal at one frequency. A periodic
signal that is not a pure sinusoid is a sum of phasors, one at each
integer multiple of a fundamental frequency. This is the
\engterm{Fourier series}, and the phasor apparatus of this section is
exactly the notation it needs. We state the construction and work one
case; the analytic machinery (convergence, the integral formulas for
the coefficients) arrives in Volume IV, and the discrete version that
a computer runs arrives in the FFT box below.

A signal $x(t)$ periodic with period $T$ and fundamental angular
frequency $\omega_{0} = 2\pi/T$ is represented by the complex Fourier
series
\[
x(t) = \sum_{n = -\infty}^{\infty} \tilde X_{n}\, e^{i n \omega_{0} t},
\qquad
\tilde X_{n} = \frac{1}{T} \int_{0}^{T} x(t)\, e^{-i n \omega_{0} t}\, dt.
\]
Each coefficient $\tilde X_{n}$ is a phasor: a complex number carrying
the amplitude and the phase of the $n$th harmonic. The signal is the
sum of those phasors, each rotating at its own frequency
$n\omega_{0}$. For a real signal the coefficients satisfy the
conjugate symmetry $\tilde X_{-n} = \overline{\tilde X_{n}}$, so the
positive and negative harmonics combine into real cosines and the
two-sided spectrum carries the same information as its positive half.
The phasor sum of a single frequency in section 3.4.2 is the Fourier
series truncated to its $n = \pm 1$ terms; the general series adds the
harmonics.

\paragraph{Worked example: the square wave as harmonic phasors.}
Take the odd square wave of amplitude one and period $T$, equal to
$+1$ on the first half-period and $-1$ on the second. The coefficient
integral gives $\tilde X_{n} = 0$ for even $n$ (the wave has half-wave
symmetry, which kills every even harmonic) and, for odd $n$,
\[
\tilde X_{n} = \frac{1}{T} \int_{0}^{T} x(t)\, e^{-i n \omega_{0} t}\, dt
= \frac{2}{i n \pi}, \qquad n \text{ odd}.
\]
Combining the $+n$ and $-n$ terms, the real time-domain series is
\[
x(t) = \frac{4}{\pi} \sum_{n = 1, 3, 5, \dots} \frac{1}{n}\sin(n\omega_{0} t)
= \frac{4}{\pi}\left[\sin\omega_{0} t + \tfrac{1}{3}\sin 3\omega_{0} t
+ \tfrac{1}{5}\sin 5\omega_{0} t + \cdots\right].
\]
The harmonic amplitudes fall off as $1/n$, the slow decay that a
sharp-edged waveform demands. Figure~\ref{fig:vol02:ch03:fourier-square}
draws the partial sums and the magnitude spectrum. The partial sum
overshoots near each discontinuity by about nine percent of the step
height no matter how many harmonics are kept, the \engterm{Gibbs
phenomenon}; the overshoot narrows as $N$ grows but does not shrink,
which is the warning that a truncated Fourier series of a discontinuous
signal never converges uniformly. The working consequence for the
engineer is that a square-wave drive into a linear network excites
every odd harmonic, and a filter that passes the fundamental but not
the harmonics recovers a sinusoid from the square wave, the operating
principle of the simplest sine-wave oscillator built from a comparator
and a band-pass filter \cite[chapter 4]{hist:fourier1822}.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-fourier-square.tex}

The reason the Fourier series belongs in a phasor chapter is
operational. A linear time-invariant network responds to each
harmonic phasor independently, scaling it by the transfer function
$H(n\omega_{0})$ at that harmonic's frequency. The network's response
to the whole periodic signal is the sum of the responses to the
individual harmonic phasors, recombined. Phasor analysis at a single
frequency, repeated once per harmonic and summed, is the complete
steady-state analysis of a linear network driven by any periodic
signal. The superposition that licenses the recombination is the same
linearity that licensed adding two phasors at one frequency in
section 3.4.2.

\begin{mastery}
\textbf{From Fourier series to the FFT.} The Fourier series
represents a continuous periodic signal by an infinite set of harmonic
phasors. A computer holds neither a continuous signal nor an infinite
set; it holds $N$ samples and computes $N$ phasors. The
\engterm{discrete Fourier transform} (DFT) of $N$ samples
$x_{0}, \dots, x_{N-1}$ is
\[
\tilde X_{k} = \sum_{n = 0}^{N-1} x_{n}\, e^{-i 2\pi k n / N},
\qquad k = 0, 1, \dots, N-1.
\]
The factor $e^{-i 2\pi k n / N}$ is the $n$th power of the $N$th root
of unity $w_{N} = e^{-i 2\pi / N}$ from section 3.3, evaluated at
harmonic $k$. The DFT is therefore a sum over the roots of unity, and
the $N$ output phasors are the discrete harmonics of the sampled
signal. Computed directly, the DFT costs $N^{2}$ complex
multiplications. The \engterm{fast Fourier transform} (FFT) factors
the $N$-point transform into two $N/2$-point transforms recursively,
exploiting the symmetry $w_{N}^{k + N/2} = -w_{N}^{k}$ of the roots of
unity, and reduces the cost to $N\log_{2} N$. For $N = 2^{20}$ samples
the speedup is a factor of about fifty thousand, the difference
between a transform that runs in milliseconds and one that runs in
minutes. The FFT is the most-executed numerical algorithm in
engineering practice: it runs inside every spectrum analyser, every
audio codec, every digital communications receiver, and every
medical-imaging reconstruction. Its mathematical content is the
roots-of-unity geometry of section 3.3 and the harmonic-phasor
decomposition of this section, made discrete and made fast
\cite[chapter 9]{text:oppenheim-schafer}.
\end{mastery}

\section{Rotations in 2D}\pathtag{standard}

A unit complex number $e^{i\theta}$ multiplies any complex number
it meets by rotating it through angle $\theta$ around the origin.
This connects the complex-multiplication rule to the geometry of
two-dimensional rotations. We make the connection explicit and
show how the same rotation is represented as a $2\times 2$ matrix
acting on the real plane.

\subsection{Multiplication by a unit complex number is rotation}

A complex number $z = r e^{i\alpha}$ has magnitude $r$ and
argument $\alpha$. The product $e^{i\theta} \cdot z = r e^{i(\theta
+ \alpha)}$ has the same magnitude $r$ and argument $\alpha +
\theta$. The point $z$ in the complex plane has been rotated
counterclockwise by $\theta$ around the origin; its distance from
the origin is unchanged. A unit complex number is a rotation
operator. The complex-multiplication rule is the algebraic content
of rotation in two dimensions.

A worked example: take $z = 1 + i$ (magnitude $\sqrt{2}$, argument
$\pi/4$). Multiplying by $e^{i\pi/2} = i$ gives $i(1 + i) = -1 +
i$ (magnitude $\sqrt{2}$, argument $3\pi/4$). The rotation is
through $\pi/2$, exactly as expected. Figure~\ref{fig:vol02:ch03:rotation}
displays the picture.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-rotation.tex}

\subsection{The $2\times 2$ rotation matrix}

In real-coordinate notation, the same rotation acts on $(x, y)$
through the matrix
\[
R(\theta) = \begin{pmatrix} \cos\theta & -\sin\theta \\ \sin\theta & \cos\theta \end{pmatrix}.
\]
The reader can verify the matrix is correct by writing the
complex multiplication $(\cos\theta + i\sin\theta)(x + iy) =
(x\cos\theta - y\sin\theta) + i(x\sin\theta + y\cos\theta)$ and
reading off the real and imaginary parts as the components of
$R(\theta)\begin{pmatrix}x \\ y\end{pmatrix}$. The two
representations agree.

The rotation matrix has working properties the reader should
keep available. The product of two rotations is the rotation by
the sum of the angles: $R(\theta_{1}) R(\theta_{2}) = R(\theta_{1}
+ \theta_{2})$. The inverse of a rotation is the rotation by the
negative angle: $R(\theta)^{-1} = R(-\theta) = R(\theta)^{T}$. The
determinant of a rotation matrix is one, and its eigenvalues are
$e^{i\theta}$ and $e^{-i\theta}$, which says again that the
rotation has no real eigenvectors when $\theta$ is not a multiple
of $\pi$.

\subsection{Geometric proof of the angle-sum identities}

The angle-sum identities follow from the matrix product
$R(\alpha) R(\beta) = R(\alpha + \beta)$. Writing both sides
explicitly and equating the $(1, 1)$ and $(2, 1)$ entries:
\begin{align*}
\cos(\alpha + \beta) &= \cos\alpha\cos\beta - \sin\alpha\sin\beta, \\
\sin(\alpha + \beta) &= \sin\alpha\cos\beta + \cos\alpha\sin\beta.
\end{align*}
The matrix proof and the complex-multiplication proof are the
same proof in two notations; rotations compose, and rotations
commute (in two dimensions, although not in three or higher), and
the algebraic content of the composition is the angle-sum
identity.

\subsection{Reflections and the orthogonal group}

A reflection across a line through the origin at angle $\theta/2$
to the real axis is the matrix
\[
F(\theta) = \begin{pmatrix} \cos\theta & \sin\theta \\ \sin\theta & -\cos\theta \end{pmatrix}.
\]
The determinant of $F(\theta)$ is $-1$, and $F(\theta)^{2} = I$.
Reflections and rotations together form the group of orthogonal
$2\times 2$ matrices, $O(2)$; rotations alone form the special
orthogonal group $SO(2)$. The reader meets these groups in formal
linear algebra (Volume II Chapter 9) and in the symmetry analyses
of crystallography and structural mechanics. We name them here so
the reader recognises the structure when it appears.

\subsection{Worked example: rotating a coordinate frame}

A strain gauge rosette measures normal strain along three axes at
$0^{\circ}$, $45^{\circ}$, and $90^{\circ}$ to a reference
direction. Converting between the gauge frame and the principal-axis
frame is a planar rotation. We take a simpler instance to fix the
mechanics. A point sits at $(x, y) = (3, 1)$ in a measurement frame.
The analysis frame is rotated counterclockwise by $30^{\circ}$
relative to the measurement frame; we want the point's coordinates
in the analysis frame.

A point fixed in space, viewed from a frame rotated by $+\theta$,
has coordinates rotated by $-\theta$. The coordinate transformation
is therefore $R(-30^{\circ})$ applied to $(3, 1)$:
\[
\begin{pmatrix} x' \\ y' \end{pmatrix}
= \begin{pmatrix} \cos(-30^{\circ}) & -\sin(-30^{\circ}) \\
\sin(-30^{\circ}) & \cos(-30^{\circ}) \end{pmatrix}
\begin{pmatrix} 3 \\ 1 \end{pmatrix}
= \begin{pmatrix} \tfrac{\sqrt 3}{2} & \tfrac{1}{2} \\
-\tfrac{1}{2} & \tfrac{\sqrt 3}{2} \end{pmatrix}
\begin{pmatrix} 3 \\ 1 \end{pmatrix}.
\]
Carrying out the multiplication, $x' = \tfrac{3\sqrt 3}{2} +
\tfrac{1}{2} = \tfrac{3\sqrt 3 + 1}{2} \approx 3.098$ and $y' =
-\tfrac{3}{2} + \tfrac{\sqrt 3}{2} = \tfrac{\sqrt 3 - 3}{2} \approx
-0.634$. The complex-number route does the same work in one
multiplication: the point is $z = 3 + i$, the frame rotation by
$+30^{\circ}$ means multiplying the point's coordinates by
$e^{-i30^{\circ}}$, and $z' = (3 + i)e^{-i\pi/6} = (3 +
i)(\tfrac{\sqrt 3}{2} - \tfrac{i}{2})$, whose real and imaginary
parts reproduce $x'$ and $y'$. The two routes agree because the
$2\times 2$ rotation matrix is the matrix of multiplication by
$e^{i\theta}$. The sign of the rotation is the recurring trap: a
frame rotated by $+\theta$ transforms coordinates by $-\theta$, and
confusing the two flips the off-diagonal signs. The geometric
reading, multiplication by a unit complex number rotates the plane,
is the discipline that keeps the sign straight
\cite[chapter 1]{text:v2ch03-needham-visual-complex}.

\section{Spherical and cylindrical coordinates preview}\pathtag{standard}

Two coordinate systems extend the polar idea to three dimensions.
The cylindrical system extends the polar coordinates by adding a
height; the spherical system replaces the height with a polar
angle and uses two angles instead of one. Both are working
notations in physical engineering. We define them and state their
working conversions; the calculus that operates on them, the
volume elements and the gradient and the divergence, arrives in
Chapters 8 and 11.

\subsection{Cylindrical coordinates}

A point in three-dimensional space at Cartesian coordinates
$(x, y, z)$ has \engterm{cylindrical coordinates} $(\rho, \phi,
z)$, where $\rho$ is the distance from the $z$-axis, $\phi$ is
the angle from the positive $x$-axis to the projection of the
point onto the $xy$-plane (measured counterclockwise looking down
the $z$-axis), and $z$ is the height. The conversions are
\begin{align*}
x &= \rho\cos\phi, & y &= \rho\sin\phi, & z &= z, \\
\rho &= \sqrt{x^{2} + y^{2}}, & \phi &= \operatorname{atan2}(y, x), & z &= z.
\end{align*}
The cylindrical system is natural for any system whose geometry
has rotational symmetry around an axis: a wire, a fluid in a
pipe, an antenna pattern around a tower, the field around a
coaxial cable. The radial direction $\hat{\rho}$, the azimuthal
direction $\hat{\phi}$, and the axial direction $\hat{z}$ form a
right-handed orthonormal basis at each point; the basis vectors
$\hat{\rho}$ and $\hat{\phi}$ rotate as the point moves around
the axis, which complicates the calculus when we get to it but
keeps the geometry transparent here.

\subsection{Spherical coordinates}

A point at Cartesian coordinates $(x, y, z)$ has \engterm{spherical
coordinates} $(r, \theta, \phi)$, where $r$ is the distance from
the origin, $\theta$ is the polar angle (the angle from the
positive $z$-axis to the line from the origin to the point,
measured in the range $[0, \pi]$), and $\phi$ is the azimuthal
angle (the same as in the cylindrical system). The conversions
are
\begin{align*}
x &= r\sin\theta\cos\phi, & y &= r\sin\theta\sin\phi, & z &= r\cos\theta, \\
r &= \sqrt{x^{2} + y^{2} + z^{2}}, & \theta &= \arccos(z/r), & \phi &= \operatorname{atan2}(y, x).
\end{align*}
The spherical system is natural for any system whose geometry has
rotational symmetry around a point: the field of a point charge,
the gravitational field of a planet, the radiation pattern of an
antenna in the far field, the seismic wavefront from a point
source.

\paragraph{Worked example: a point in three coordinate systems.}
Take the Cartesian point $(x, y, z) = (1, 1, \sqrt{2})$ and express
it in cylindrical and spherical coordinates. The cylindrical radius
is $\rho = \sqrt{1 + 1} = \sqrt{2}$; the azimuth is $\phi =
\operatorname{atan2}(1, 1) = \pi/4$; the height is $z = \sqrt{2}$.
So the cylindrical coordinates are $(\sqrt{2}, \pi/4, \sqrt{2})$. For
the spherical coordinates, the distance from the origin is $r =
\sqrt{1 + 1 + 2} = 2$; the azimuth is the same $\phi = \pi/4$; the
polar angle is $\theta = \arccos(z/r) = \arccos(\sqrt{2}/2) = \pi/4$.
So the spherical coordinates are $(2, \pi/4, \pi/4)$. The point sits
at $45^{\circ}$ from the $z$-axis and at $45^{\circ}$ azimuth, a
symmetric position that the spherical coordinates make legible and
the Cartesian coordinates obscure. A reader who reverses the
conversion, computing $x = r\sin\theta\cos\phi = 2 \cdot
\tfrac{\sqrt 2}{2} \cdot \tfrac{\sqrt 2}{2} = 1$, recovers the
original point as a check. The azimuth $\phi$ is the same number in
both systems and is recovered by the same four-quadrant arctangent,
which is the section 3.1 inverse-tangent caution carried into three
dimensions.

A working caution. The convention $(r, \theta, \phi)$ used here,
with $\theta$ as polar angle and $\phi$ as azimuthal angle, is
common in physics. A common engineering convention swaps the
names: $(r, \phi, \theta)$ with $\phi$ as polar angle and $\theta$
as azimuthal angle. A third convention, common in geography, uses
latitude (measured from the equator, in the range $[-\pi/2,
\pi/2]$) instead of the polar angle (measured from the pole). The
reader meeting a spherical-coordinate formula should check the
convention before substituting; an angle measured from the pole
and an angle measured from the equator are complementary, and a
sine and cosine are exchanged.

\subsection{When each system is right}

The choice between Cartesian, cylindrical, and spherical
coordinates is dictated by the symmetry of the problem. A
calculation that ignores the symmetry, by reaching for Cartesian
coordinates because they are the most familiar, often does the
work three or six times instead of once. A calculation that uses
the symmetry, by reaching for cylindrical or spherical coordinates
when the geometry has axial or point symmetry, often produces a
clean answer in one or two lines. We meet the apparatus that uses
the systems in Volume II Chapters 8 and 11; the reader who has
internalised the conversions here will read those chapters as
physics rather than as bookkeeping.

\section{Failure: phase wrap-around in real measurements}\pathtag{core}

Phase is an angle defined modulo $2\pi$. A measurement of phase
that does not track the cycle count loses the difference between
$\theta$ and $\theta + 2\pi$, $\theta + 4\pi$, $\theta - 2\pi$,
and so on indefinitely. The failure mode is straightforward to
describe and is responsible for a recurring class of measurement
errors in working systems. We name the mechanism, give the
typical symptoms, and describe the discipline that keeps the
mechanism contained.

\subsection{The mechanism}

A phase meter, a phase-locked loop, an interferometer, a
synchrophasor, a radar Doppler processor, or any instrument that
reports a phase angle returns a value in a finite range, almost
always $(-\pi, \pi]$ or $[0, 2\pi)$. The instrument computes the
phase of the sampled signal at the instant of measurement; it
does not generally count the integer number of cycles that have
elapsed since some reference instant. When the underlying
physical phase advances past $\pi$, the reported value
discontinuously jumps to $-\pi$ and continues to advance from
there. The reported phase is the \engterm{wrapped phase}; the
underlying physical phase, including the cycle count, is the
\engterm{unwrapped phase}.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-phase-wrap.tex}

A phase-velocity calculation that depends on the unwrapped phase
will fail at every wrap point. A range measurement by phase-shift
ranging (which uses the phase difference between a transmitted
and a received signal to infer distance) becomes ambiguous beyond
half a wavelength. A vibration measurement that integrates a
phase-derivative signal accumulates an offset of $2\pi$ at every
wrap that is not detected. The pathology has the same shape in
every domain: the working signal is the unwrapped phase, the
measured signal is the wrapped phase, and the engineer who treats
the wrapped phase as the working signal will report a velocity
that drops to zero at every wrap or a position that snaps back at
every cycle.

\subsection{Symptoms in working measurements}

The reader meets phase wrap-around as a measurement defect in at
least the following common settings.

\textit{Phase-shift ranging.} An instrument transmits a sinusoidal
signal of wavelength $\lambda$ and receives the reflected signal
from a target. The phase difference between transmitted and
received signals is proportional to twice the range, modulo
$2\pi$. The instrument can report the range modulo $\lambda/2$
unambiguously, but cannot resolve which integer number of
half-wavelengths separates the instrument from the target. A
laser interferometer with $\lambda = 633 \,\si{\nano\meter}$ has
a wrap distance of $\sim 316 \,\si{\nano\meter}$; the working
discipline is to start from a known reference position and
track the unwrapped phase by counting wraps as the target moves.

\textit{Synchrophasors in power systems.} A phasor measurement
unit reports the phase of the local mains voltage relative to a
GPS-synchronised reference at, typically, $30$ to $60$ samples
per second (current as of 2026). The wrapped phase is reported on
$(-\pi, \pi]$. A network operator computing the phase difference
between two ends of a long transmission line must unwrap the
reported phase before computing power flow; a wrap that is not
recognised produces a spurious power-flow excursion of order
gigawatts on a continental grid. The discipline is to track the
phase trajectory between samples and add or subtract $2\pi$
whenever the reported phase jumps by more than a chosen threshold
(typically $\pi$).

\textit{Doppler radar and sonar.} A coherent radar measures the
phase of the returned signal at successive pulses; the
pulse-to-pulse phase change is proportional to the target's
radial velocity times the pulse interval. A target moving fast
enough that the inter-pulse phase change exceeds $\pi$ produces
an aliased velocity report; the radar's \engterm{unambiguous
velocity range} is set by the pulse repetition frequency. The
discipline is to choose the pulse repetition frequency to bracket
the expected velocity range, or to use staggered pulse intervals
that disambiguate the aliasing.

\textit{Numerical phase computations.} A program that computes
phase as $\arctan(\Im(z)/\Re(z))$ instead of
$\operatorname{atan2}(\Im(z), \Re(z))$ collapses the four
quadrants of the complex plane into two and introduces a
$\pi$-radian error for points in the second and third quadrants.
The bug is silent: the program returns a number; the number is
just wrong by $\pi$. The discipline is to use the four-quadrant
arctangent in every phase computation and to test the
quadrant-boundary cases.
Figure~\ref{fig:vol02:ch03:quadrant-signs} shows the mechanism: a
point in the second quadrant is mapped by the naive $\arctan$ onto a
direction in the fourth quadrant, an error of exactly $\pi$.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-quadrant-signs.tex}

\textit{Phase noise and wrap clustering.} When a signal's phase
is near a wrap boundary and the noise is comparable to the
wrap range, successive samples may report phases that bounce
across the boundary. A simple unwrap algorithm that thresholds at
$\pi$ may cumulatively drift by $\pm 2\pi$ as it
chases the noise. The discipline is to filter the noisy phase
before unwrapping, or to use an unwrap algorithm that considers
more than the immediate sample-to-sample change (a smoothing
filter, a Kalman estimator, or a path-following algorithm).

\subsection{A second failure: small-angle approximation past its range}

The wrap failure is the chapter's principal failure mode. A second,
quieter failure shares its shape: a trigonometric approximation used
outside the range where it holds. The small-angle approximations
$\sin\theta \approx \theta$ and $\tan\theta \approx \theta$
degrade smoothly with the angle, and a calculation that adopts them
at $5^{\circ}$ and silently carries them to $30^{\circ}$ accumulates
an error that no single line of the calculation flags.

The pendulum is the canonical case. The small-angle pendulum has a
period $T = 2\pi\sqrt{L/g}$ that is independent of amplitude, the
property that made the pendulum the timekeeping standard for two and
a half centuries. The exact period grows with amplitude through a
correction $T = 2\pi\sqrt{L/g}\,[1 + \tfrac{1}{16}\theta_{0}^{2} +
\cdots]$, where $\theta_{0}$ is the swing amplitude. At a $10^{\circ}$
amplitude ($\theta_{0} \approx 0.175\,\text{rad}$) the correction is
about $0.19\%$, which over a day is some $170\,\si{\second}$ of
accumulated error for a clock regulated on the small-angle period.
The precision pendulum clock holds its amplitude small and constant
for exactly this reason. An engineer who reuses the small-angle
period at a $30^{\circ}$ amplitude, where the correction has grown to
about $1.7\%$, has imported a systematic error of order one part in
sixty into every prediction.

The discipline mirrors the wrap discipline. State the angle range
over which the approximation is adopted. State the error the
approximation incurs at the edge of that range. Verify, at least
once, against the exact expression. Flag any input that pushes the
angle past the stated range. The approximation is a tool with a
specified operating envelope, and using it outside the envelope is
the same category of error as treating a wrapped phase as an
unwrapped one: a model applied past the boundary where it holds.

\subsection{The unwrap discipline}

A reliable phase measurement that depends on the unwrapped phase
takes care of the wrapping at every stage. The working
discipline has four parts.

First, identify whether the application depends on the unwrapped
phase or only on the wrapped phase. Power-factor measurement at a
single instant uses only the wrapped phase. Range measurement,
velocity measurement, and any time-integrated phase signal use
the unwrapped phase.

Second, choose the sampling rate or the pulse rate so that the
phase advances by less than $\pi$ between samples. The criterion
is the temporal version of the Nyquist condition; the analogue
of the Nyquist frequency is the rate at which phase wraps occur
in the underlying physical signal.

Third, run an unwrap algorithm on the sampled phase that adds
$2\pi k$ to each sample so that consecutive samples differ by less
than $\pi$. A first-order unwrap algorithm examines the
difference between successive samples and adjusts by $\pm 2\pi$
when the difference exceeds $\pi$ in magnitude. Higher-order
algorithms (smoothing, filtering, Bayesian estimators) are
available when the noise is comparable to the wrap range; the
literature on phase unwrapping in synthetic-aperture radar is the
working reference \cite{text:v2ch03-ghiglia-pritt-unwrap}.

Fourth, verify the unwrap against an independent check: the
total time elapsed and the expected phase advance, the residual
range against an absolute reference, the cumulative cycle count
recorded separately by a coarser sensor. A phase measurement
that has not been verified end-to-end against an independent
reference is a phase measurement whose unwrap could be wrong.

\begin{principle}[The wrap discipline]
A measurement that reports a phase angle reports the wrapped
phase. A calculation that depends on the unwrapped phase must
include an explicit unwrap step, an explicit verification of the
unwrap against an independent reference, and an explicit check
that the sampling rate is high enough to keep the phase advance
between samples below $\pi$ radians. A program that omits the
$\operatorname{atan2}$ four-quadrant convention is broken in the
second and third quadrants. A range measurement by phase-shift
ranging is ambiguous beyond half a wavelength without an
independent coarse range.
\end{principle}

\section{Worked examples: AC circuit, rotating machinery, wave addition}\pathtag{core}

We close the chapter's apparatus with three worked examples that
exercise the trigonometric, the complex-number, and the phasor
machinery on the working applications of section 3.4. Each example
runs the full calculation; each is followed by a one-line summary
of what the example demonstrated.

\subsection{Series RLC circuit at a fixed frequency}

A series circuit consists of a resistor $R$, an inductor $L$,
and a capacitor $C$ connected in series across a sinusoidal
voltage source $v_{s}(t) = V_{m}\cos(\omega t)$. We compute the
steady-state current and the voltage across each element.

The phasor of the source is $\tilde{V}_{s} = V_{m}$ (taking
$\varphi = 0$ as the reference phase). The series impedance is
\[
Z = Z_{R} + Z_{L} + Z_{C} = R + i\omega L - \frac{i}{\omega C} = R + i\left(\omega L - \frac{1}{\omega C}\right).
\]
The phasor of the current is $\tilde{I} = \tilde{V}_{s}/Z$. We
write $Z = |Z| e^{i\varphi_{Z}}$ with
\[
|Z| = \sqrt{R^{2} + \left(\omega L - \frac{1}{\omega C}\right)^{2}}, \qquad \varphi_{Z} = \arctan\!\left(\frac{\omega L - 1/(\omega C)}{R}\right).
\]
Then $\tilde{I} = (V_{m}/|Z|) e^{-i\varphi_{Z}}$, so the
time-domain current is
\[
i(t) = \frac{V_{m}}{|Z|}\cos(\omega t - \varphi_{Z}).
\]

A working numerical example. Take $R = 100 \,\si{\ohm}$,
$L = 100 \,\si{\milli\henry}$, $C = 10 \,\si{\micro\farad}$,
$V_{m} = 10 \,\si{\volt}$, and $\omega = 2\pi \cdot 50 \,\si{\hertz}
\approx 314 \,\si{\radian\per\second}$. Then $\omega L \approx
31.4 \,\si{\ohm}$, $1/(\omega C) \approx 318 \,\si{\ohm}$, and
$\omega L - 1/(\omega C) \approx -287 \,\si{\ohm}$. The
impedance magnitude is $|Z| \approx \sqrt{100^{2} + 287^{2}}
\approx 304 \,\si{\ohm}$. The phase angle is $\varphi_{Z}
\approx \arctan(-287/100) \approx -1.236 \,\text{rad} \approx
-70.8^{\circ}$. The current amplitude is $|V_{m}|/|Z| \approx
0.0329 \,\si{\ampere}$, and the current leads the source by
$70.8^{\circ}$ (because $\varphi_{Z}$ is negative, $-\varphi_{Z}$
is positive, and the current's phase is the negative of the
impedance's phase angle).

The voltage across each element is
\begin{align*}
\tilde{V}_{R} &= R \tilde{I} \approx (100)(0.0329) e^{i \cdot 70.8^{\circ}} \,\si{\volt}, \\
\tilde{V}_{L} &= i\omega L \tilde{I} \approx (31.4)(0.0329) e^{i(70.8^{\circ} + 90^{\circ})} \,\si{\volt}, \\
\tilde{V}_{C} &= -\frac{i}{\omega C} \tilde{I} \approx (318)(0.0329) e^{i(70.8^{\circ} - 90^{\circ})} \,\si{\volt},
\end{align*}
with magnitudes approximately $3.29 \,\si{\volt}$, $1.03 \,\si
{\volt}$, and $10.46 \,\si{\volt}$ respectively. The capacitor
voltage exceeds the source voltage in magnitude; this is a
common feature of below-resonance operation in a series RLC
circuit and not an arithmetic error. The reader can verify by
adding the three element voltages as phasors and obtaining the
source phasor: $\tilde{V}_{R} + \tilde{V}_{L} + \tilde{V}_{C} =
\tilde{V}_{s}$, which is Kirchhoff's voltage law in the phasor
form. The numerical values cross-check against the standard
worked treatment of the series RLC branch in
\cite[chapter 8]{text:v2ch03-irwin-nelms}.

The example demonstrates the steps of phasor analysis on the
canonical RLC circuit. Figure~\ref{fig:vol02:ch03:phasor-rlc}
displays the phasor diagram and the corresponding time-domain
waveforms; the dashed return arrow on the phasor diagram is the
geometric statement of Kirchhoff's voltage law. The chapter
project asks the reader to repeat the calculation for a
parameter set of their choice and to write a reflection on what
the phasor representation buys over time-domain integration.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-phasor-rlc.tex}

\subsection{Case study: a two-branch network on a $230$-volt mains}

The series RLC example exercises one loop. A working AC analysis
usually has more than one loop, and the case study that follows runs
the full phasor procedure on a two-branch network from setup to
interpretation, with every number carried to the end. The network,
drawn in figure~\ref{fig:vol02:ch03:ac-network}, is a deliberately
ordinary one: a small single-phase load that an electrician would meet
behind a wall plate. A motor-and-cable branch (a resistance in series
with an inductance) and a power-factor-correction capacitor in
parallel with a bleeder resistance share one node and return to the
supply. We compute the branch currents, the supply current, the power
delivered, and the power factor, and we read what the capacitor buys.

\paragraph{The setup.} The supply is a $230\,\si{\volt}$ RMS,
$50\,\si{\hertz}$ single-phase mains (the European domestic standard,
current as of 2026). We take the supply voltage as the phase reference
and use the peak-value phasor convention of section 3.4, so the source
phasor is $\tilde V_{s} = V_{m} = 230\sqrt{2} \approx 325.3\,\si{\volt}$
at angle zero. The angular frequency is $\omega = 2\pi \cdot 50 \approx
314.2\,\si{\radian\per\second}$. The two branches are:
\begin{itemize}[leftmargin=*]
  \item \emph{Branch 1 (motor-and-cable)}: a resistance $R_{1} =
    20\,\si{\ohm}$ in series with an inductance $L = 60\,\si{\milli
    \henry}$. This branch models a small induction motor and its
    feeder cable, the inductance standing in for the motor's
    magnetising reactance.
  \item \emph{Branch 2 (correction)}: a capacitance $C = 40\,\si
    {\micro\farad}$ in parallel with a bleeder resistance $R_{2} =
    2000\,\si{\ohm}$ that discharges the capacitor when the supply is
    removed.
\end{itemize}
The two branches connect across the same supply node $\mathrm{N}$ and
the return, so they are in parallel as seen by the source.

\paragraph{Step one: branch impedances.} Branch 1 has inductive
reactance $X_{L} = \omega L = 314.2 \cdot 0.060 \approx 18.85\,\si
{\ohm}$, so
\[
Z_{1} = R_{1} + i X_{L} = 20 + 18.85\,i\,\si{\ohm},
\]
with magnitude $|Z_{1}| = \sqrt{20^{2} + 18.85^{2}} \approx 27.48\,\si
{\ohm}$ and angle $\arg Z_{1} = \arctan(18.85/20) \approx 0.7558\,
\text{rad} \approx 43.3^{\circ}$. Branch 2 has capacitive reactance
$X_{C} = 1/(\omega C) = 1/(314.2 \cdot 40 \times 10^{-6}) \approx
79.58\,\si{\ohm}$, so its capacitor impedance is $-79.58\,i\,\si
{\ohm}$, in parallel with $R_{2} = 2000\,\si{\ohm}$:
\[
Z_{2} = \frac{R_{2} \cdot (-i X_{C})}{R_{2} - i X_{C}}
= \frac{2000 \cdot (-79.58\,i)}{2000 - 79.58\,i}.
\]
The denominator has magnitude $\sqrt{2000^{2} + 79.58^{2}} \approx
2001.6\,\si{\ohm}$ and angle $\arctan(-79.58/2000) \approx -0.03978\,
\text{rad}$. The numerator is $-159{,}160\,i = 159{,}160\,e^{-i\pi/2}\,
\si{\ohm}^{2}$. Dividing, $Z_{2} \approx (159{,}160/2001.6)\,
e^{i(-\pi/2 + 0.03978)} \approx 79.51\,e^{-i1.531}\,\si{\ohm}$, which
in rectangular form is $Z_{2} \approx 3.16 - 79.45\,i\,\si{\ohm}$. The
$2\,\si{\kilo\ohm}$ bleeder is large enough that branch 2 is very
nearly a pure capacitor; its small real part is the bleeder's residual
dissipation.

\paragraph{Step two: branch currents.} Each branch sees the full
supply voltage. By Ohm's law in phasor form,
\[
\tilde I_{1} = \frac{\tilde V_{s}}{Z_{1}}
= \frac{325.3}{27.48\,e^{i 0.7558}}
\approx 11.84\,e^{-i 0.7558}\,\si{\ampere},
\]
a current of peak amplitude $11.84\,\si{\ampere}$ ($8.37\,\si{\ampere}$
RMS) lagging the supply by $43.3^{\circ}$, the inductive branch. For
branch 2,
\[
\tilde I_{2} = \frac{\tilde V_{s}}{Z_{2}}
= \frac{325.3}{79.51\,e^{-i 1.531}}
\approx 4.09\,e^{i 1.531}\,\si{\ampere},
\]
a current of peak amplitude $4.09\,\si{\ampere}$ ($2.89\,\si{\ampere}$
RMS) leading the supply by $87.7^{\circ}$, the capacitive branch.

\paragraph{Step three: the supply current.} Kirchhoff's current law
at node $\mathrm{N}$ gives the supply current as the sum of the branch
currents. Converting each to rectangular form before adding (the rule
of section 3.3: add in rectangular, multiply in polar):
\begin{align*}
\tilde I_{1} &\approx 11.84(\cos(-0.7558) + i\sin(-0.7558))
  \approx 8.62 - 8.12\,i\,\si{\ampere}, \\
\tilde I_{2} &\approx 4.09(\cos(1.531) + i\sin(1.531))
  \approx 0.164 + 4.09\,i\,\si{\ampere}.
\end{align*}
The sum is
\[
\tilde I = \tilde I_{1} + \tilde I_{2}
\approx (8.62 + 0.164) + i(-8.12 + 4.09)
= 8.78 - 4.03\,i\,\si{\ampere}.
\]
The supply current has peak magnitude $|\tilde I| = \sqrt{8.78^{2} +
4.03^{2}} \approx 9.66\,\si{\ampere}$ ($6.83\,\si{\ampere}$ RMS) and
angle $\arg \tilde I = \arctan(-4.03/8.78) \approx -0.4308\,\text{rad}
\approx -24.7^{\circ}$. The capacitor's leading current has cancelled
much of the inductive branch's lagging current: the supply now sees a
current that lags by only $24.7^{\circ}$ instead of the
motor branch's $43.3^{\circ}$.

\paragraph{Step four: power and power factor.} The complex power
delivered by the supply, in the peak convention, is $S = \tfrac{1}{2}
\tilde V_{s}\bar{\tilde I}$:
\[
S = \tfrac{1}{2}(325.3)(8.78 + 4.03\,i)
\approx 1428 + 656\,i\,\text{VA}.
\]
The average power delivered is $P = \Re(S) \approx 1428\,\si{\watt}$;
the reactive power is $Q = \Im(S) \approx 656\,\text{var}$; the apparent
power is $|S| = \sqrt{1428^{2} + 656^{2}} \approx 1572\,\text{VA}$.
The power factor is $P/|S| \approx 1428/1572 \approx 0.908$ lagging.
A check: the real power must all be dissipated in the two resistances.
The RMS branch currents are $8.37\,\si{\ampere}$ and $2.89\,\si
{\ampere}$, but only the part of branch 2's current that flows in
$R_{2}$ dissipates there; the resistive dissipation is $I_{1,\text{RMS}}^{2}
R_{1} + (V_{\text{RMS}}^{2}/R_{2}) = 8.37^{2}\cdot 20 + 230^{2}/2000
\approx 1401 + 26.5 \approx 1428\,\si{\watt}$, which matches $\Re(S)$
to the rounding. The power balance closes.

\paragraph{Interpretation: what the capacitor bought.} Without the
correction capacitor, the supply would carry only branch 1's current,
$8.37\,\si{\ampere}$ RMS at a power factor of $\cos(43.3^{\circ})
\approx 0.728$ lagging, an apparent power of $230 \cdot 8.37 \approx
1925\,\text{VA}$ for the same $1401\,\si{\watt}$ of real motor
power. The capacitor raises the power factor from $0.728$ to $0.908$
and cuts the supply current from $8.37\,\si{\ampere}$ to $6.83\,\si
{\ampere}$ RMS, a $19\%$ reduction in the current the cable and the
upstream transformer must carry, for the same useful work. The cost is
the capacitor and its bleeder, and the small extra dissipation in the
bleeder. The whole calculation is the four-step phasor procedure of
section 3.4.7 run on a two-node network: replace each element by its
impedance, write Kirchhoff's laws as phasor algebra, solve, and read
off the time-domain quantities. A time-domain solution would have set
up a coupled pair of first-order differential equations and integrated
them past the transient; the phasor calculation is a page of complex
arithmetic. The numerical values cross-check against the standard
worked treatment of parallel AC branches and power-factor correction
in \cite[chapter 10]{text:v2ch03-nilsson-riedel}. A reader who reaches
for a calculator should confirm the supply current and power factor
before reading further; the supply current $\tilde I$ is the quantity
that sizes the breaker, and a sign error in the capacitive branch
flips the correction into a worsening.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-ac-network.tex}

\subsection{Rotating machinery and the polyphase representation}

A three-phase electrical machine has three windings displaced by
$120^{\circ}$ ($2\pi/3$ radians) around the machine's axis. When
driven by a balanced three-phase supply, the three winding
voltages are
\begin{align*}
v_{a}(t) &= V_{m}\cos(\omega t), \\
v_{b}(t) &= V_{m}\cos(\omega t - 2\pi/3), \\
v_{c}(t) &= V_{m}\cos(\omega t - 4\pi/3).
\end{align*}
The phasor representations are $\tilde{V}_{a} = V_{m}$,
$\tilde{V}_{b} = V_{m} e^{-i2\pi/3}$, $\tilde{V}_{c} = V_{m}
e^{-i4\pi/3}$. The three phasors are at angles $0$, $-120^{\circ}$,
$-240^{\circ}$ (equivalently, $0$, $-120^{\circ}$, $+120^{\circ}$)
and they sum to zero:
\[
\tilde{V}_{a} + \tilde{V}_{b} + \tilde{V}_{c} = V_{m}\left(1 + e^{-i2\pi/3} + e^{-i4\pi/3}\right) = 0,
\]
because the three phasors are the cube roots of $V_{m}^{3}$ and
the sum of the cube roots of any positive real number is zero.
This algebraic identity is the engineering reason that a balanced
three-phase system needs no neutral conductor: the phase currents
sum to zero at every instant in steady state.
Figure~\ref{fig:vol02:ch03:three-phase} draws the three phasors as a
star and the three waveforms whose instantaneous sum is zero.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-three-phase.tex}

A working numerical check. Take $V_{m} = 1$ for simplicity. The
three phasors are $\tilde V_{a} = 1$, $\tilde V_{b} = e^{-i2\pi/3} =
-\tfrac{1}{2} - i\tfrac{\sqrt 3}{2}$, and $\tilde V_{c} =
e^{-i4\pi/3} = e^{i2\pi/3} = -\tfrac{1}{2} + i\tfrac{\sqrt 3}{2}$.
Adding: the real parts give $1 - \tfrac{1}{2} - \tfrac{1}{2} = 0$,
and the imaginary parts give $0 - \tfrac{\sqrt 3}{2} + \tfrac{\sqrt
3}{2} = 0$. The sum is exactly zero, as the roots-of-unity identity
promised. The same cancellation holds at every instant in the time
domain: $\cos\omega t + \cos(\omega t - 2\pi/3) + \cos(\omega t +
2\pi/3) = 0$ for all $t$, which the reader can verify by the
sum-to-product identity applied to the second and third terms.

When the machine is driven in such a way that the three currents
are also balanced and shifted from their respective voltages by a
common angle, the magnetic field inside the machine rotates at
constant angular velocity $\omega$. The rotating field is the
working principle of the synchronous and induction motor and the
generator. Its constant magnitude follows from the same identity
that gives the cube roots of unity: the squared magnitudes of the
three current contributions add to a constant, and the rotating
field's magnitude is the square root of that constant. The
rotation preserves structure; the archetype recurs.

\begin{archetype}[Stability: the rotating field of constant magnitude]
The rotating magnetic field of a balanced polyphase machine is the
stability archetype in its most consequential engineering form. Three
sinusoidal currents, each pulsating in its own winding, combine into
a field vector that rotates at constant angular velocity with constant
magnitude. The conserved quantity is the field magnitude; the varying
quantity is the field angle. The same structure that the unit circle
displays (a radius held fixed while the angle advances) appears here
as torque held steady while the rotor turns. A single-phase machine
has a pulsating field whose magnitude varies through the cycle, which
produces a torque ripple at twice the line frequency and a starting
problem; the polyphase machine trades one winding for two or three
and buys a field whose magnitude is conserved over the cycle. The
engineer recognises the archetype every time a coordinate
distinguishes ``where in the rotation'' from ``how much,'' and
arranges the system so that the ``how much'' is the conserved
quantity.
\end{archetype}

\subsection{Wave addition}

Two waves of the same frequency travelling in the same direction
combine into a single wave of the same frequency. The amplitude
and phase of the combined wave are read off the sum of the two
phasors. Take two waves
\[
y_{1}(x, t) = A_{1}\cos(kx - \omega t + \varphi_{1}), \qquad y_{2}(x, t) = A_{2}\cos(kx - \omega t + \varphi_{2}).
\]
At a fixed point $x$, the time-domain phasors are $\tilde{Y}_{1}
= A_{1} e^{i\varphi_{1}}$ and $\tilde{Y}_{2} = A_{2} e^{i\varphi_{2}}$;
the combined phasor is the sum, with magnitude
\[
|\tilde{Y}_{1} + \tilde{Y}_{2}| = \sqrt{A_{1}^{2} + A_{2}^{2} + 2 A_{1} A_{2}\cos(\varphi_{2} - \varphi_{1})}.
\]
The familiar limits are visible. When $\varphi_{2} - \varphi_{1}
= 0$, the combined amplitude is $A_{1} + A_{2}$ (constructive
interference). When $\varphi_{2} - \varphi_{1} = \pi$, the
combined amplitude is $|A_{1} - A_{2}|$ (destructive interference).
At intermediate phases the combined amplitude lies between these
extremes; the formula is the law of cosines applied to the
triangle whose two sides are $A_{1}$ and $A_{2}$ and whose
included angle is $\pi - (\varphi_{2} - \varphi_{1})$.

The same calculation underlies the double-slit interference
pattern, the two-element antenna array's directional pattern, the
combined gravitational wave from a binary inspiral seen from two
detectors, and the audio interference between two loudspeakers in
a room. The phase difference $\varphi_{2} - \varphi_{1}$ is the
working quantity that controls whether two signals reinforce or
cancel, and the phasor representation reads off the answer in one
line.

A working numerical example: two loudspeakers driven in phase emit
sound at $1\,\si{\kilo\hertz}$. A listener stands so that the path
length from the second speaker is $0.17\,\si{\meter}$ longer than
from the first. The wavelength at $1\,\si{\kilo\hertz}$ in air
(speed $\approx 343\,\si{\meter\per\second}$, current as of 2026) is
$\lambda = 343/1000 = 0.343\,\si{\meter}$. The extra path introduces
a phase difference $\Delta\varphi = 2\pi\,(0.17)/(0.343) \approx
3.11\,\text{rad} \approx 178^{\circ}$, close to $\pi$. The two
contributions, taken equal in amplitude $A$, combine to
$\sqrt{A^{2} + A^{2} + 2A^{2}\cos(178^{\circ})} = A\sqrt{2 +
2\cos(178^{\circ})} \approx A\sqrt{2 - 1.999} \approx 0.035 A$. The
listener sits in a near-null: the two speakers nearly cancel because
the half-metre path difference is close to half a wavelength. Moving
$0.17\,\si{\meter}$ to bring the path difference to a full wavelength
restores constructive interference and a near-doubling of amplitude.
The phasor sum reads off the answer in one line, and the result is
the everyday experience of comb-filtered dead spots in a room with
two sources.

\paragraph{Worked example: beating between two close frequencies.}
The same-frequency phasor sum fails the moment the two signals differ
in frequency, because a phasor is defined only at one frequency: two
phasors at different frequencies rotate at different rates and their
sum is not a phasor at all. The right tool is the sum-to-product
identity of section 3.1. Take two tones of equal amplitude $A$ at
frequencies $f_{1} = 440\,\si{\hertz}$ and $f_{2} = 444\,\si{\hertz}$,
the situation of a musician tuning a second string against a reference
$A$. Their sum is
\[
A\cos(2\pi f_{1} t) + A\cos(2\pi f_{2} t)
= 2A\cos\!\left(2\pi \tfrac{f_{1} + f_{2}}{2} t\right)
\cos\!\left(2\pi \tfrac{f_{2} - f_{1}}{2} t\right).
\]
The first factor is a fast carrier at the average frequency $(f_{1} +
f_{2})/2 = 442\,\si{\hertz}$; the second factor is a slow envelope at
half the difference frequency, $(f_{2} - f_{1})/2 = 2\,\si{\hertz}$.
The ear hears the carrier as a single pitch at $442\,\si{\hertz}$ whose
loudness swells and fades. The loudness follows $|\cos(2\pi \cdot 2 t)|$,
whose period is half the envelope's period because the rectification
of hearing folds the negative half of the envelope onto the positive,
so the audible \engterm{beat frequency} is the full difference
$f_{2} - f_{1} = 4\,\si{\hertz}$: four loudness pulses per second. The
musician tightens the string until the four-per-second beat slows to
zero, at which point the two frequencies match. The beat is a
high-resolution null detector: a four-hertz beat against a
$440\,\si{\hertz}$ reference is a frequency error of under one percent,
heard plainly where a direct pitch comparison would not resolve it.
The same construction is the working principle of the heterodyne
frequency counter, the optical beat note that compares two lasers, and
the Doppler-shift difference tone in radar; the difference frequency is
the measurand, recovered as the envelope rate.

\subsection{Application sketches: where the apparatus actually
runs}

The phasor and complex-number apparatus carries the working weight
of several common engineering subsystems. We sketch four, each in
two paragraphs, so the reader meets the application at the right
scale without having to wait until the volume that owns it.

\textit{Synchrophasors and grid-wide state estimation.}
A phasor measurement unit (PMU) reports the magnitude and phase of
the local mains voltage and current against a GPS-locked timing
reference, at $30$ or $60$ frames per second under the IEEE
C37.118 standard \cite{paper:v2ch03-itrich-c37-118}. The phase is
the quantity that ties a measurement at one substation to a
measurement at another substation hundreds of kilometres away. A
transmission system operator that aggregates synchrophasor reports
from across a continental grid runs a complex-valued state
estimator whose unknowns are voltage phasors at every node; the
estimator's residuals flag bad data and locate emerging
instabilities minutes before they would otherwise become visible.

The apparatus depends on the wrap discipline of section 3.7. Phase
is reported on the principal range $(-\pi, \pi]$. Computing power
flow on a long line ($P = (|V_{1}||V_{2}|/X)\sin(\theta_{1} -
\theta_{2})$ on a lossless inductive line, current as of 2026)
requires the unwrapped phase difference, recovered by tracking the
phase trajectory between samples and adding or subtracting $2\pi$
at every wrap. A missed wrap on a $300\,\si{\kilo\meter}$
transmission line operating near its surge impedance loading shifts
the inferred power flow by hundreds of megawatts; the
state-estimation residual is the operational guard that flags the
error before any control action is taken.

\textit{Heterodyne and lock-in detection.} A lock-in amplifier
multiplies a noisy input signal by a reference signal at a known
frequency and phase. The product-to-sum identity from section 3.1
shows that the multiplication produces content at the sum and the
difference of the two frequencies; a low-pass filter selects the
difference frequency (zero if the input and reference are at the
same frequency), leaving a slowly-varying signal proportional to
the in-phase or the quadrature component of the input at the
reference frequency. The instrument extracts a small signal buried
in noise over many decades of dynamic range; the working
specification of a research-grade lock-in amplifier is a noise
floor in the nanovolt range with a few-hertz output bandwidth.

The same product-to-sum identity is the operating principle of the
superheterodyne radio receiver. An incoming radio-frequency
signal at carrier frequency $f_{c}$ is multiplied by a local
oscillator at $f_{c} - f_{IF}$ to produce a difference signal at
the intermediate frequency $f_{IF}$, which can then be filtered and
demodulated by circuitry whose tuning does not vary as the
receiver is retuned to a different carrier. Every radio receiver
manufactured since the 1930s is a working implementation of the
heterodyne identity.

\textit{Three-phase rotating-field machines.} The polyphase
identity of section 3.8 (three balanced cosines spaced by
$120^{\circ}$ sum to zero at every instant in steady state) is the
operating principle behind the synchronous generator, the
induction motor, and the three-phase rectifier. A balanced
three-phase supply, applied to a stator with three windings spaced
$120^{\circ}$ around the rotor axis, produces a magnetic field of
constant magnitude that rotates at the supply frequency. The rotor
sees a rotating field rather than a pulsating field; the torque
is smooth rather than twice-line-frequency oscillating, and the
machine starts on its own without an auxiliary commutator.

The same identity, run backwards, gives the three-phase generator:
a rotor mechanically driven at the synchronous speed and excited
with direct current induces three sinusoidal voltages in the
stator windings spaced $120^{\circ}$ apart, which combine to
deliver constant instantaneous power to a balanced load. The
generators at every utility power plant in the world produce
three-phase power for exactly this reason: the rotating field's
constant-magnitude property is the architectural feature that
keeps the power flow steady at the source-load interface.

\textit{Optical and microwave interferometers.} A Michelson
interferometer divides a coherent beam between two arms, reflects
each back to a beam splitter, and combines the two returns onto a
detector. The detector sees the magnitude squared of the sum of
the two complex amplitudes; by section 3.8 (wave addition), the
intensity is
$I = I_{1} + I_{2} + 2\sqrt{I_{1} I_{2}}\cos(\Delta\varphi)$,
where $\Delta\varphi$ is the round-trip phase difference between
the two arms. A path-length change of half a wavelength shifts
$\Delta\varphi$ by $2\pi$ and cycles the intensity through one
fringe. The LIGO gravitational-wave detector is a kilometre-scale
Michelson interferometer that measures path-length changes of
order $10^{-18}\,\si{\meter}$ by tracking the fractional fringe
shift; the readout chain is built around the wrap discipline of
section 3.7 because each detected event covers many fringes and
the unwrapped phase is the working signal.

\section{The complex plane as a working instrument}\pathtag{standard}

The phasor reduces a single-frequency steady state to one complex
number. The complex plane carries more than that. Where a phasor lives
on the plane (which quadrant, how close to an axis, how far from the
origin) reads off as physical behaviour, and the geometry of points,
circles, and distances becomes a working instrument the engineer reaches
for without re-deriving the algebra each time. We work five applications
that exercise the plane this way: a symmetrical-components fault, an
impedance-matching network read on the Smith chart, an inverse transform
by residues, a filter response read off its poles and zeros, and an
antenna pattern as a phasor sum. Each runs end to end; each leaves the
reader a procedure rather than a result.

\subsection{Case study: a single-line-to-ground fault by symmetrical
components}

The two-branch mains network of section 3.8 ran the phasor procedure on
a balanced single-phase load. Three-phase faults are the setting where
the complex plane earns its keep a second way, through a change of basis
that turns a coupled three-phase problem into three uncoupled
single-phase problems. We work a single-line-to-ground fault on a small
distribution feeder from the sequence networks to the fault current, the
calculation a protection engineer runs to size a breaker and set a
relay.

\paragraph{The decomposition.} A three-phase set of phasors
$(\tilde V_{a}, \tilde V_{b}, \tilde V_{c})$ need not be balanced. A
broken conductor, an unbalanced load, or a fault destroys the symmetry,
and the per-phase calculation that worked for the balanced supply no
longer reduces to a single phase. Fortescue's 1918 construction resolves
any three-phase set into three balanced sets: a \engterm{positive
sequence} with the original $a$-$b$-$c$ rotation, a \engterm{negative
sequence} with the reversed $a$-$c$-$b$ rotation, and a \engterm{zero
sequence} of three equal in-phase phasors. Writing $\alpha = e^{i 2\pi/3}$
for the rotation by $120^{\circ}$ (a cube root of unity from
section 3.3), the decomposition and its inverse are
\[
\begin{pmatrix} \tilde V_{a} \\ \tilde V_{b} \\ \tilde V_{c} \end{pmatrix}
= \begin{pmatrix} 1 & 1 & 1 \\ 1 & \alpha^{2} & \alpha \\ 1 & \alpha & \alpha^{2} \end{pmatrix}
\begin{pmatrix} \tilde V_{0} \\ \tilde V_{1} \\ \tilde V_{2} \end{pmatrix},
\qquad
\begin{pmatrix} \tilde V_{0} \\ \tilde V_{1} \\ \tilde V_{2} \end{pmatrix}
= \frac{1}{3}\begin{pmatrix} 1 & 1 & 1 \\ 1 & \alpha & \alpha^{2} \\ 1 & \alpha^{2} & \alpha \end{pmatrix}
\begin{pmatrix} \tilde V_{a} \\ \tilde V_{b} \\ \tilde V_{c} \end{pmatrix},
\]
with subscripts $0$, $1$, $2$ for zero, positive, and negative sequence.
The matrix is built from the cube roots of unity; the factor $1/3$ and
the conjugate arrangement make the inverse the conjugate-transpose up to
scale, the same orthogonality the discrete Fourier transform of length
three would show.
Figure~\ref{fig:vol02:ch03:symmetrical-components} draws the three
sequence stars.

The change of basis pays off because the sequence components do not
interact in a balanced power system. A symmetric three-phase line has
the same impedance to positive-sequence current as to negative-sequence
current, and a separate (usually larger) impedance to zero-sequence
current, which must return through ground or a neutral. The three
sequence networks are uncoupled, so a fault analysis solves three
single-phase circuits and recombines.

\paragraph{The fault.} Take a feeder fed from a source whose
positive-sequence voltage behind its impedance is $\tilde V_{f} =
1.0\,\text{pu}$ (per-unit, the dimensionless ratio to a nominal base, the
working unit of power-system analysis). The sequence impedances seen from
the fault point, in per-unit on the same base, are
\[
Z_{1} = 0.02 + 0.20\,i, \qquad
Z_{2} = 0.02 + 0.20\,i, \qquad
Z_{0} = 0.05 + 0.60\,i,
\]
typical for a small distribution transformer and its feeder (current as
of 2026). Phase $a$ faults to ground through negligible fault impedance.

The boundary conditions at the fault are the physical content. Phase $a$
is shorted to ground, so $\tilde V_{a} = 0$; phases $b$ and $c$ carry no
fault current, so $\tilde I_{b} = \tilde I_{c} = 0$. Pushing these
through the inverse transform, the three sequence currents are equal:
\[
\tilde I_{0} = \tilde I_{1} = \tilde I_{2}
= \frac{\tilde V_{f}}{Z_{0} + Z_{1} + Z_{2}}.
\]
The equal-sequence-currents result is the algebraic reason the three
sequence networks connect in series for a single-line-to-ground fault.
The reader who memorises one fault connection memorises this one; the
others (line-to-line, double-line-to-ground) follow from their own
boundary conditions.

\paragraph{The numbers.} Sum the sequence impedances:
\[
Z_{0} + Z_{1} + Z_{2} = (0.05 + 0.02 + 0.02) + i(0.60 + 0.20 + 0.20)
= 0.09 + 1.00\,i\,\text{pu}.
\]
The magnitude is $|Z_{\text{sum}}| = \sqrt{0.09^{2} + 1.00^{2}} \approx
1.004\,\text{pu}$ and the angle is $\arctan(1.00/0.09) \approx
1.481\,\text{rad} \approx 84.9^{\circ}$, a nearly pure reactance as
faults usually are. Each sequence current is
\[
\tilde I_{1} = \frac{1.0}{1.004\,e^{i 1.481}} \approx 0.996\,e^{-i 1.481}\,\text{pu},
\]
lagging the driving voltage by $84.9^{\circ}$. The fault current in the
faulted phase is the sum of the three equal sequence currents:
\[
\tilde I_{a} = \tilde I_{0} + \tilde I_{1} + \tilde I_{2} = 3\,\tilde I_{1}
\approx 2.99\,e^{-i 1.481}\,\text{pu}.
\]
The factor of three is the headline result: a bolted single-line-to-ground
fault on this feeder draws about three per-unit of current, three times
the nominal load current. If the base current is $400\,\si{\ampere}$, the
fault current is about $1.2\,\si{\kilo\ampere}$, and the breaker must
interrupt that current within its rated clearing time while the relay
must see it as a fault and not as an inrush.

\paragraph{What the basis change bought.} A direct three-phase solution
would have written a coupled $6 \times 6$ real system (three complex
phase voltages, three complex phase currents, with the mutual coupling
between phases filling the off-diagonal). The sequence transform
diagonalises the coupling, because the symmetric line is
\engterm{circulant} in the phase index and the cube roots of unity are
its eigenvectors. The fault analysis becomes three scalar divisions and
one sum. The same eigenstructure that makes the balanced three-phase
currents sum to zero in section 3.8 makes the sequence networks
uncouple here; both are the roots-of-unity geometry of section 3.3 read
in a power-system frame. The standard reference treatment of sequence
networks and fault types is in \cite[chapter 11]{text:v2ch03-nilsson-riedel},
and the synchrophasor-based protection that runs on top of it is the
C37.118 apparatus of section 3.8 \cite{paper:v2ch03-itrich-c37-118}.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-symmetrical-components.tex}

\subsection{Impedance matching and the Smith chart}

The conformal-map mastery box named the Smith chart as the image of the
impedance plane under the Mobius transformation. We work the chart as an
instrument: match a complex load to a real source with a two-element
network, reading each step as motion on the chart.

\paragraph{The reflection coefficient.} A load impedance $Z_{L}$
connected to a line of characteristic impedance $Z_{0}$ reflects a
fraction of an incident wave back toward the source. The fraction is the
\engterm{reflection coefficient}
\[
\Gamma = \frac{Z_{L} - Z_{0}}{Z_{L} + Z_{0}}
= \frac{z - 1}{z + 1}, \qquad z = \frac{Z_{L}}{Z_{0}},
\]
where $z$ is the impedance normalised to $Z_{0}$. The map $z \mapsto
\Gamma$ is a Mobius transformation. It sends the impedance right-half-plane
(all passive loads, with non-negative resistance) onto the closed unit
disk $|\Gamma| \le 1$, because a passive load cannot reflect more power
than arrives. The matched condition $Z_{L} = Z_{0}$ maps to $\Gamma = 0$,
the centre of the disk; a short ($z = 0$) maps to $\Gamma = -1$ and an
open ($z \to \infty$) to $\Gamma = +1$. The Smith chart is the picture of
this disk with the images of the constant-resistance and
constant-reactance lines drawn in, as in
figure~\ref{fig:vol02:ch03:smith-chart}.

\paragraph{The matching task.} A load $Z_{L} = 25 - 30\,i\,\si{\ohm}$ is
to be matched to a $Z_{0} = 50\,\si{\ohm}$ source at $f = 100\,\si
{\mega\hertz}$. Normalised, $z = 0.5 - 0.6\,i$. The reflection coefficient
is
\[
\Gamma = \frac{(0.5 - 0.6\,i) - 1}{(0.5 - 0.6\,i) + 1}
= \frac{-0.5 - 0.6\,i}{1.5 - 0.6\,i}.
\]
Rationalise by $1.5 + 0.6\,i$: the numerator is $(-0.5 - 0.6\,i)(1.5 +
0.6\,i) = -0.75 - 0.30\,i - 0.90\,i - 0.36\,i^{2} = -0.39 - 1.20\,i$, and
the denominator is $1.5^{2} + 0.6^{2} = 2.61$. So $\Gamma \approx -0.149 -
0.460\,i$, with magnitude $|\Gamma| \approx 0.483$. The matching network
must move this point to the origin.

\paragraph{The L-network.} The standard two-element match adds a shunt
susceptance and then a series reactance (or the reverse), each chosen so
the working point glides along a constant-conductance circle and then a
constant-resistance circle to the centre. Working in admittance for the
shunt element, the load admittance is
\[
Y_{L} = \frac{1}{Z_{L}} = \frac{1}{25 - 30\,i}
= \frac{25 + 30\,i}{25^{2} + 30^{2}}
= \frac{25 + 30\,i}{1525}
\approx (16.4 + 19.7\,i)\,\si{\milli\siemens}.
\]
A shunt element of susceptance $B$ adds to the imaginary part of $Y_{L}$.
Choosing $B$ to land the conductance-circle intersection where the series
step can finish the match, the design adds a shunt inductor that
contributes $B_{L} = -1/(\omega L_{\text{sh}})$ to cancel part of the
load susceptance and bring the real part of the resulting impedance to
$Z_{0}$. Carrying the algebra (the closed-form L-match solution for
$R_{L} < Z_{0}$) gives a shunt susceptance near $-13.5\,\si{\milli
\siemens}$ and a remaining series reactance near $+27\,\si{\ohm}$, which
at $100\,\si{\mega\hertz}$ is a shunt inductor of about
$L_{\text{sh}} = 1/(\omega |B|) \approx 118\,\si{\nano\henry}$ and a
series inductor of about $L_{\text{se}} = X/\omega \approx 43\,\si
{\nano\henry}$. The exact component values depend on which of the two
L-network orientations is chosen; both land the working point at the
centre of the chart, and both are read off as a two-segment path from
the load point to $\Gamma = 0$.

\paragraph{The reading.} The match removes the reflection: at
$\Gamma = 0$, all incident power enters the load and none returns. The
\engterm{voltage standing-wave ratio}, $\text{VSWR} = (1 + |\Gamma|)/(1 -
|\Gamma|)$, drops from $(1 + 0.483)/(1 - 0.483) \approx 2.87$ before the
match to $1.00$ after it. The Smith chart turns the algebra of the Mobius
map into a drafting exercise; the engineer who internalises that
constant-resistance circles map to circles and that adding a series
reactance slides the point along one of them does the match with a
compass and a straightedge. The full chart construction and the
two-element and three-element matching networks are the working content
of \cite[chapter 2]{text:pozar-microwave}; the conformal-mapping
geometry underneath is in \cite[chapter 6]{text:v2ch03-needham-visual-complex}.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-smith-chart.tex}

\subsection{Inverse transforms by residues and partial fractions}

The Laplace mastery box of section 3.4 replaced $i\omega$ with a complex
$s$ and named the poles of a transfer function as the points where it
fails to be analytic. Recovering the time-domain response from the
transform is a complex-plane operation: a sum of residues at those poles,
or equivalently a partial-fraction split followed by term-by-term
inversion. We work both routes on one transfer function so the reader
sees they are the same calculation.

\paragraph{The transform.} A step voltage applied to a series RLC circuit
produces a capacitor-voltage transform of the form
\[
V_{C}(s) = \frac{\omega_{0}^{2}}{s\,(s^{2} + 2\sigma s + \omega_{0}^{2})},
\]
with $\omega_{0} = 1/\sqrt{LC}$ the undamped natural frequency and
$\sigma = R/(2L)$ the damping rate. Take an underdamped case with
$\omega_{0} = 1000\,\si{\radian\per\second}$ and $\sigma = 200\,\si
{\radian\per\second}$. The damped natural frequency is $\omega_{d} =
\sqrt{\omega_{0}^{2} - \sigma^{2}} = \sqrt{10^{6} - 4 \times 10^{4}} =
\sqrt{9.6 \times 10^{5}} \approx 980\,\si{\radian\per\second}$. The
denominator factors as $s\,(s - p)(s - \bar p)$ with the conjugate pole
pair
\[
p = -\sigma + i\omega_{d} \approx -200 + 980\,i, \qquad
\bar p \approx -200 - 980\,i,
\]
and a real pole at the origin from the step.

\paragraph{Route one: partial fractions.} Write
\[
V_{C}(s) = \frac{A}{s} + \frac{B}{s - p} + \frac{\bar B}{s - \bar p},
\]
where the conjugate symmetry of the pole pair forces the residues at $p$
and $\bar p$ to be conjugate. The residue at the origin is $A = \lim_{s
\to 0} s\,V_{C}(s) = \omega_{0}^{2}/(p\,\bar p) = \omega_{0}^{2}/
|p|^{2} = \omega_{0}^{2}/\omega_{0}^{2} = 1$, using $|p|^{2} = \sigma^{2}
+ \omega_{d}^{2} = \omega_{0}^{2}$. The residue at $p$ is
\[
B = \lim_{s \to p}(s - p)\,V_{C}(s)
= \frac{\omega_{0}^{2}}{p\,(p - \bar p)}
= \frac{\omega_{0}^{2}}{p \cdot 2 i\omega_{d}}.
\]
With $p = -\sigma + i\omega_{d}$ of magnitude $\omega_{0}$ and argument
$\theta_{p} = \pi - \arctan(\omega_{d}/\sigma)$, the residue is
\[
B = \frac{\omega_{0}^{2}}{2\omega_{0}\omega_{d}}\,e^{-i(\theta_{p} + \pi/2)}
= \frac{\omega_{0}}{2\omega_{d}}\,e^{-i(\theta_{p} + \pi/2)}.
\]
The number $B$ is one complex residue; its conjugate covers the other
pole.

\paragraph{Route two: residues directly.} The inverse Laplace transform
is the Bromwich contour integral, and the residue theorem evaluates it as
the sum of residues of $V_{C}(s)\,e^{st}$ over all poles to the left of
the contour:
\[
v_{C}(t) = \sum_{\text{poles}} \operatorname{Res}\left[V_{C}(s)\,e^{st}\right].
\]
The residue at the origin is $A\,e^{0 \cdot t} = 1$. The residues at $p$
and $\bar p$ are $B\,e^{pt}$ and $\bar B\,e^{\bar p t}$, a conjugate pair
whose sum is twice the real part of either:
\[
B\,e^{pt} + \bar B\,e^{\bar p t} = 2\,\Re\!\left[B\,e^{pt}\right]
= 2\,|B|\,e^{-\sigma t}\cos(\omega_{d} t + \arg B).
\]
Assembling, with $|B| = \omega_{0}/(2\omega_{d})$,
\[
v_{C}(t) = 1 + \frac{\omega_{0}}{\omega_{d}}\,e^{-\sigma t}
\cos(\omega_{d} t + \arg B), \qquad t \ge 0.
\]
The step response is unity (the final capacitor voltage) plus a decaying
oscillation at the damped frequency $\omega_{d}$, with envelope
$e^{-\sigma t}$ set by the pole's distance from the imaginary axis and
phase $\arg B$ fixed by the residue. The two routes give the same answer
because a partial-fraction coefficient is a residue; the residue calculus
is the bookkeeping that pairs each pole with one exponential mode in the
time domain. The reader meets the formal contour integral and the residue
theorem in a complex-analysis course \cite[chapter 3]{text:v2ch03-stein-shakarchi-complex};
the working route for inverting a rational transform is the
partial-fraction split, with the conjugate-pole shortcut $2\,\Re[\cdot]$
doing the work the angle-sum identities would otherwise do.

\subsection{A filter response read off the complex plane}

A transfer function $H(s)$ of a linear circuit is a rational function of
$s$: a ratio of polynomials whose roots are the \engterm{zeros}
(numerator roots) and \engterm{poles} (denominator roots). The steady-state
frequency response is $H(i\omega)$, the transfer function evaluated on the
imaginary axis. The magnitude and phase at each frequency are read off
the geometry of the pole-zero plot without evaluating a single
trigonometric function.

\paragraph{The geometric rule.} Write $H(s) = K\,\prod_{m}(s - z_{m}) /
\prod_{n}(s - p_{n})$. At the test point $s = i\omega$ on the imaginary
axis,
\[
|H(i\omega)| = |K|\,\frac{\prod_{m}|i\omega - z_{m}|}{\prod_{n}|i\omega - p_{n}|},
\qquad
\arg H(i\omega) = \arg K + \sum_{m}\arg(i\omega - z_{m})
- \sum_{n}\arg(i\omega - p_{n}).
\]
Each factor $|i\omega - z_{m}|$ is the distance from the test point to a
zero, and each $|i\omega - p_{n}|$ is the distance to a pole. The
magnitude is the product of distances to the zeros divided by the product
of distances to the poles. The phase is the sum of the angles the zero
vectors make minus the sum of the angles the pole vectors make. The
response is read by sliding the test point up the imaginary axis and
watching the distances.

\paragraph{The band-pass example.} A second-order band-pass filter has
the transfer function $H(s) = K s / [(s - p)(s - \bar p)]$ with a zero at
the origin and a conjugate pole pair $p = -\sigma_{0} + i\omega_{0}$ in
the left half-plane. Take $\omega_{0} = 1000\,\si{\radian\per\second}$ and
$\sigma_{0} = 100\,\si{\radian\per\second}$, a quality factor $Q =
\omega_{0}/(2\sigma_{0}) = 5$. At low frequency ($\omega \ll \omega_{0}$)
the zero at the origin is close to the test point, so the numerator
distance is small and the magnitude is suppressed: the filter rejects low
frequencies. At high frequency ($\omega \gg \omega_{0}$) both pole
distances grow without bound while the zero distance grows only linearly,
so the magnitude falls off: the filter rejects high frequencies. Near
$\omega = \omega_{0}$ the test point passes closest to the upper pole, the
near-pole distance shrinks to its minimum value $\sigma_{0}$, and the
magnitude peaks. The peak height is set by how small $\sigma_{0}$ is: a
pole close to the imaginary axis (small damping, high $Q$) gives a tall,
narrow peak; a pole far from the axis gives a low, broad one.
Figure~\ref{fig:vol02:ch03:pole-zero-bode} draws the pole-zero plot and
the magnitude response side by side.

\paragraph{The reading as a design tool.} A filter designer places poles
and zeros to shape $|H(i\omega)|$, then realises the placement with
circuit elements. The bandwidth is twice the pole's distance from the
imaginary axis: $\Delta\omega = 2\sigma_{0}$. The centre frequency is the
pole's distance from the real axis: $\omega_{0}$. Moving the pole toward
the imaginary axis sharpens the filter and, at the limit $\sigma_{0} \to
0$, puts the pole on the axis and makes the filter an undamped
oscillator. A pole that crosses into the right half-plane ($\sigma_{0} <
0$) makes the response grow without bound: the system is unstable, the
$e^{-\sigma_{0} t}$ envelope of the residue calculation becoming a
growing $e^{+|\sigma_{0}| t}$. The left-half-plane condition for stability
is the time-domain decay condition read on the complex plane, the bridge
to the control material of Volume V. The pole-zero reading is the working
method behind \cite[chapter 14]{text:alexander-sadiku-circuits} and the
classical filter tables of \cite[chapter 2]{text:ch06t3-williams-electronic-filter-2006}.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-pole-zero-bode.tex}

\subsection{An antenna array as a phasor sum}

The wave-addition formula of section 3.8 added two phasors. An antenna
array adds many, and the geometry of the sum is a directional pattern.
The same roots-of-unity cancellation that zeroes a balanced three-phase
current shapes the array's nulls.

\paragraph{The array factor.} A \engterm{uniform linear array} of $N$
identical elements spaced a distance $d$ apart, fed with equal amplitude
and a progressive phase increment $\beta$ from element to element, radiates
a field in direction $\vartheta$ that is the sum of $N$ phasors. Element
$n$ contributes a phasor $e^{i n\psi}$, where the \engterm{electrical
angle}
\[
\psi = k d \cos\vartheta + \beta, \qquad k = \frac{2\pi}{\lambda},
\]
collects the path-length phase $k d \cos\vartheta$ from the inter-element
spacing and the feed phase $\beta$. The total is a geometric series:
\[
AF(\psi) = \sum_{n = 0}^{N-1} e^{i n\psi}
= \frac{e^{i N\psi} - 1}{e^{i\psi} - 1}
= e^{i(N-1)\psi/2}\,\frac{\sin(N\psi/2)}{\sin(\psi/2)}.
\]
The factoring uses the same $e^{i\theta} - 1 = 2 i\,e^{i\theta/2}
\sin(\theta/2)$ identity that the Euler formulas of section 3.3 supply.
The magnitude is the \engterm{array factor}
\[
|AF(\psi)| = \left|\frac{\sin(N\psi/2)}{\sin(\psi/2)}\right|,
\]
the Dirichlet kernel, drawn in
figure~\ref{fig:vol02:ch03:array-factor}.

\paragraph{The reading.} At $\psi = 0$ all $N$ phasors point the same way
and add coherently to $N$: the main lobe. The array factor falls to zero
whenever $N\psi/2$ is a multiple of $\pi$ but $\psi/2$ is not, that is, at
$\psi = 2\pi m / N$ for $m = 1, \ldots, N-1$. At each such $\psi$ the $N$
phasors are the $N$th roots of unity rotated, spaced evenly around the
circle, and their sum is exactly zero by the roots-of-unity identity of
section 3.3. The $N - 1$ nulls between main lobes are the angles at which
the phasor polygon closes. The first null sets the beamwidth; the depth
of the nulls sets the array's ability to reject an interferer from a
specific direction.

\paragraph{The numbers.} A six-element array ($N = 6$) at half-wavelength
spacing ($d = \lambda/2$, so $kd = \pi$) fed in phase ($\beta = 0$) has
$\psi = \pi\cos\vartheta$. The main lobe is at $\vartheta = 90^{\circ}$
(broadside, $\psi = 0$). The first null is at $\psi = 2\pi/6 = \pi/3$,
which is $\cos\vartheta = 1/3$, that is, $\vartheta \approx 70.5^{\circ}$
and its mirror at $109.5^{\circ}$: a main-lobe half-width of about
$19.5^{\circ}$. Steering the beam off broadside means choosing $\beta \ne
0$ to shift the $\psi = 0$ point to a non-zero $\vartheta$; a
\engterm{phased array} steers electronically by varying $\beta$ with no
moving parts, which is the operating principle of modern radar and of the
beamforming in a multi-antenna wireless base station. The full array
theory, including the element pattern that multiplies the array factor
and the grating lobes that appear when $d > \lambda/2$, is in
\cite[chapter 6]{text:balanis-antennas}; the phasor-sum core is the
geometric series above, read on the unit circle.

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-array-factor.tex}

\begin{mastery}
\textbf{Branch cuts and Riemann surfaces.} The $n$th-root construction of
section 3.3 returned $n$ distinct values, and the argument was defined
only modulo $2\pi$. These facts have a consequence for complex functions
that the engineer meets the moment a square root or a logarithm appears
in a transfer function. The function $w = \sqrt{z}$ cannot be made
single-valued and continuous on the whole punctured plane: following $z$
once around the origin, the argument increases by $2\pi$ and $\sqrt{z} =
\sqrt{r}\,e^{i\theta/2}$ picks up a factor $e^{i\pi} = -1$, returning to
the opposite sign. To make the square root a genuine function one cuts the
plane along a \engterm{branch cut}, conventionally the negative real axis,
across which the function is not continued; the cut plane carries one
single-valued \engterm{branch}. Stacking the two branches and gluing them
along the cut builds the \engterm{Riemann surface}, a two-sheeted cover of
the plane on which $\sqrt{z}$ is single-valued and analytic everywhere
except the \engterm{branch point} at the origin. The logarithm has
infinitely many branches, one per cycle of the argument, and an
infinite-sheeted surface. The engineering consequence is concrete: the
characteristic impedance of a transmission line, $Z_{0} = \sqrt{(R +
i\omega L)/(G + i\omega C)}$, and the propagation constant $\gamma =
\sqrt{(R + i\omega L)(G + i\omega C)}$ each carry a square root, and a
numerical routine that picks the wrong branch returns an impedance with
negative real part, a passive line that appears to generate power. The
discipline is to choose the branch that keeps the physical quantity in its
correct half-plane and to track the branch explicitly across any frequency
sweep that encircles a branch point \cite[chapter 7]{text:v2ch03-needham-visual-complex}.
\end{mastery}

\begin{mastery}
\textbf{The Laplace, Fourier, and $Z$ bridge.} Three transforms recur
across the engineering volumes, and they are one construction read on
three contours of the complex plane. The \engterm{Laplace transform}
$F(s) = \int_{0}^{\infty} f(t)\,e^{-st}\,dt$ uses the general complex
frequency $s = \sigma + i\omega$ and lives on the whole $s$-plane; its
region of convergence is a right half-plane, and its poles encode the
system's natural modes. Restricting to the imaginary axis, $s = i\omega$,
gives the \engterm{Fourier transform} $F(i\omega) = \int_{-\infty}^{\infty}
f(t)\,e^{-i\omega t}\,dt$, the steady-state frequency response, the
contour on which the phasor of this chapter lives. Sampling the
continuous signal at interval $T$ and mapping $s$ through $z = e^{sT}$
gives the \engterm{$Z$ transform} $F(z) = \sum_{n} f_{n}\,z^{-n}$, the
discrete-time analogue; the map $z = e^{sT}$ wraps the $s$-plane's
imaginary axis onto the unit circle $|z| = 1$ and folds the left
half-plane into the disk interior, which is why the stability boundary is
the imaginary axis for Laplace and the unit circle for the $Z$ transform.
The exponential map $z = e^{sT}$ is the same wrapping that makes the
sampled spectrum periodic and that produces aliasing: frequencies
differing by a multiple of the sampling rate $2\pi/T$ map to the same
point on the unit circle. The three transforms share the residue calculus
for inversion, the pole-zero reading for the response, and the
left-half-plane-or-inside-the-disk condition for stability. The reader
meets Laplace in Volume V's control material, the Fourier transform in
Volume IV's wave chapters, and the $Z$ transform in the digital-signal
chapters; the unifying picture is that each is the bilateral exponential
transform evaluated on its own contour, and the phasor is the single-point
restriction to one frequency on the Fourier contour
\cite[chapter 9]{text:oppenheim-schafer}.
\end{mastery}

\section{Further worked problems and two applications}\pathtag{standard}

The apparatus of the chapter rewards a second pass at working scale. We
extend the trigonometry beyond the right triangle, carry the complex
exponential into the logarithm and the general power, read resonance and
bandwidth off the parallel tank, and close with two applications that run
the machinery end to end on a rotating machine and on a direction-finding
array. Each worked example is self-contained; each application is two to
three pages of one concrete problem.

\subsection{Oblique triangles: the law of sines and the law of cosines}

The right-triangle ratios of section 3.1 answer questions about triangles
with a $90^{\circ}$ angle. Most triangles a surveyor, a machinist, or a
structural analyst meets have no right angle. Two identities extend the
trigonometry to the general triangle with sides $a$, $b$, $c$ opposite
angles $\alpha$, $\beta$, $\gamma$. The \engterm{law of sines} states
\[
\frac{a}{\sin\alpha} = \frac{b}{\sin\beta} = \frac{c}{\sin\gamma} = 2R_{c},
\]
where $R_{c}$ is the radius of the triangle's circumscribed circle. The
\engterm{law of cosines} states
\[
c^{2} = a^{2} + b^{2} - 2ab\cos\gamma,
\]
with the two companions obtained by cycling the labels. The law of cosines
reduces to the Pythagorean theorem when $\gamma = \pi/2$, because
$\cos(\pi/2) = 0$; the right triangle is the special case in which the
correction term vanishes. Both identities follow from dropping an altitude
and applying the right-triangle ratios to the two triangles it creates, a
derivation the reader can reconstruct and one that shows the oblique case
built from right-triangle pieces.
Figure~\ref{fig:vol02ch03:oblique-triangle} sets the two identities in a
surveying resection.

The choice between the two identities is dictated by what is known. Two
angles and any side fix the triangle through the law of sines. Two sides and
the included angle, or all three sides, fix it through the law of cosines.
The law of sines carries an ambiguity when the known data are two sides and a
non-included angle (the \engterm{side-side-angle case}), because the arcsine
returns two supplementary candidate angles and both can close a valid
triangle; the working discipline is to check whether the second candidate
produces a consistent triangle before discarding it.

\paragraph{Worked example: a surveying resection.} Two survey stations $A$
and $B$ sit a measured baseline $c = 150\,\text{m}$ apart. A theodolite at
$A$ reads the angle to a distant target $C$ as $\alpha = 37^{\circ}$ above
the baseline; a theodolite at $B$ reads the angle to the same target as
$\beta = 61^{\circ}$. Find the two ranges $\overline{AC}$ and
$\overline{BC}$. The third angle is $\gamma = 180^{\circ} - 37^{\circ} -
61^{\circ} = 82^{\circ}$. The law of sines gives the range from $A$ to $C$
as the side $b$ opposite $\beta$:
\[
b = \overline{AC} = c\,\frac{\sin\beta}{\sin\gamma}
= 150\,\frac{\sin 61^{\circ}}{\sin 82^{\circ}}
\approx 150\,\frac{0.8746}{0.9903} \approx 132.5\,\text{m},
\]
and the range from $B$ to $C$ as the side $a$ opposite $\alpha$:
\[
a = \overline{BC} = c\,\frac{\sin\alpha}{\sin\gamma}
= 150\,\frac{\sin 37^{\circ}}{\sin 82^{\circ}}
\approx 150\,\frac{0.6018}{0.9903} \approx 91.1\,\text{m}.
\]
The two ranges locate $C$ from either station. A check: the largest side
faces the largest angle, and here $\gamma = 82^{\circ}$ is the largest
angle while the baseline $c = 150\,\text{m}$ is the largest side, as it must
be. The resection is the working core of triangulation, the method that
mapped continents before satellite positioning and that still fixes a
total station's position from two known control points.

\paragraph{Worked example: the third side by the law of cosines.} A
truss member $CA$ of length $b = 4.0\,\text{m}$ meets a member $CB$ of length
$a = 3.0\,\text{m}$ at a joint $C$ where the included angle is $\gamma =
110^{\circ}$. Find the distance $c$ between the far ends and the angle the
closing member makes at $A$. The law of cosines gives
\[
c^{2} = a^{2} + b^{2} - 2ab\cos\gamma
= 9 + 16 - 2(3)(4)\cos 110^{\circ}
= 25 - 24(-0.3420) \approx 33.21\,\text{m}^{2},
\]
so $c \approx 5.76\,\text{m}$. The obtuse included angle makes the closing
side longer than either leg, which the negative cosine encodes: an included
angle past $90^{\circ}$ adds to the squared length rather than subtracting.
The angle at $A$ follows from the law of sines,
$\sin\alpha = a\sin\gamma/c = 3(0.9397)/5.76 \approx 0.4894$, so
$\alpha \approx 29.3^{\circ}$. Here the side-side-angle ambiguity does not
bite, because $\alpha$ is opposite the shorter of the two members and must be
acute. The two identities together turn a triangle with any two sides and
any known angle into a fully solved figure, which is the arithmetic a frame
analysis runs at every joint.

\subsection{The complex logarithm and general powers}

The exponential $e^{z}$ is defined for every complex $z$, and section 3.3
read off its behaviour on the imaginary axis. Its inverse, the
\engterm{complex logarithm}, extends the real logarithm and carries the
multivaluedness that the argument's modulo-$2\pi$ freedom forces. For a
nonzero complex number $z = r e^{i\theta}$,
\[
\log z = \ln r + i(\theta + 2\pi k), \qquad k \in \mathbb{Z},
\]
where $\ln r$ is the ordinary real logarithm of the magnitude and the
imaginary part is the argument, defined only up to an integer number of
turns. The \engterm{principal logarithm} $\operatorname{Log} z = \ln|z| + i\operatorname{Arg} z$ picks
the argument in $(-\pi, \pi]$, one value from the infinite family. The
logarithm converts the magnitude to a real part and the phase to an
imaginary part, which is the reason a plot of gain in decibels against
phase in degrees (the Bode pair, and the Nyquist and Nichols plots that
Volume V builds on them) is a plot of the complex logarithm of the transfer
function.

A general power follows from the logarithm through $z^{w} = e^{w\operatorname{Log} z}$,
and it inherits the logarithm's multivaluedness. The construction resolves
expressions that look meaningless in the real numbers.

\paragraph{Worked example: the value of $i^{i}$.} Compute $i^{i}$. Write the
base in polar form: $i = e^{i\pi/2}$, so $\operatorname{Log} i = i\pi/2$ (taking the
principal branch). Then
\[
i^{i} = e^{i\,\operatorname{Log} i} = e^{i\,(i\pi/2)} = e^{-\pi/2} \approx 0.2079.
\]
The imaginary unit raised to the imaginary power is a real number, and a
small one. The full multivalued answer is $i^{i} = e^{-\pi/2 - 2\pi k}$ for
integer $k$, an infinite set of positive real values, one per branch of the
logarithm; the principal branch returns the middle one. The example is not a
curiosity for its own sake. Every time a numerical routine raises a complex
base to a complex power (a fractional-order filter, a branch of a transfer
function, a fractional derivative) it selects one branch of this construction,
and a routine that selects the wrong branch returns one of the other members
of the family, off by a factor $e^{-2\pi}$ or its powers.

\paragraph{Worked example: the principal square root in rectangular form.}
The polar $n$th-root construction of section 3.3 needs the argument. When a
complex number arrives in rectangular form and only the principal square
root is wanted, a direct rectangular formula avoids the arctangent. For
$z = a + bi$ with principal square root $w = u + vi$, matching real and
imaginary parts of $w^{2} = z$ gives $u^{2} - v^{2} = a$ and $2uv = b$,
which solve to
\[
u = \sqrt{\frac{|z| + a}{2}}, \qquad
v = \operatorname{sgn}(b)\,\sqrt{\frac{|z| - a}{2}},
\]
with $|z| = \sqrt{a^{2} + b^{2}}$ and the sign of $v$ taken from the sign of
$b$ so that $2uv = b$ holds. Take $z = 3 + 4i$, of magnitude $|z| = 5$. Then
$u = \sqrt{(5 + 3)/2} = \sqrt{4} = 2$ and $v = \sqrt{(5 - 3)/2} = 1$, so the
principal square root is $w = 2 + i$, which checks: $(2 + i)^{2} = 4 + 4i +
i^{2} = 3 + 4i$. The other square root is $-w = -2 - i$. Both radicands
$|z| \pm a$ are non-negative because $|z| \ge |a|$, so the formula never
takes a square root of a negative number; the sign choice for $v$ is the only
place the branch enters, and taking it from $b$ selects the principal root
with argument in $(-\pi/2, \pi/2]$.

\subsection{Resonance, quality factor, and bandwidth in the phasor picture}

The series RLC worked example of section 3.9 evaluated the impedance at one
frequency. Sweeping the frequency reveals \engterm{resonance}, the condition
$\omega L = 1/(\omega C)$ at which the inductive and capacitive reactances
cancel. At resonance the series impedance collapses to its resistance $R$,
the current reaches its maximum $V_{m}/R$, and the voltage and current are in
phase. The resonant frequency is
\[
\omega_{0} = \frac{1}{\sqrt{LC}},
\]
the same combination that section 3.10 read as the common magnitude of a
conjugate pole pair. The sharpness of the resonance is the
\engterm{quality factor} $Q = \omega_{0} L / R = 1/(\omega_{0} C R)$, the
ratio of the reactance at resonance to the resistance. A large $Q$ means a
tall, narrow resonance; a small $Q$ means a broad, shallow one. The
\engterm{bandwidth} between the two half-power frequencies, where the power
delivered falls to half its peak and the impedance phase reaches
$\pm\pi/4$, is
\[
\Delta\omega = \frac{\omega_{0}}{Q} = \frac{R}{L}.
\]
The parallel RLC tank shows the dual behaviour: its impedance peaks at
resonance rather than dipping, drawn in
figure~\ref{fig:vol02ch03:parallel-rlc-resonance}, and its $Q$ is
$R/(\omega_{0} L) = \omega_{0} C R$, the reciprocal combination.

\paragraph{Worked example: reading $Q$ and bandwidth off a tank.} A parallel
tank uses $L = 10\,\si{\milli\henry}$, $C = 100\,\si{\nano\farad}$, and a
loss resistance $R = 5\,\si{\kilo\ohm}$ across the tank. The resonant
frequency is $\omega_{0} = 1/\sqrt{LC} = 1/\sqrt{10^{-3}\cdot 10^{-7}} =
1/\sqrt{10^{-10}} = 10^{5}\,\si{\radian\per\second}$, or $f_{0} =
\omega_{0}/(2\pi) \approx 15.9\,\si{\kilo\hertz}$. The quality factor of the
parallel tank is $Q = R/(\omega_{0} L) = 5000/(10^{5}\cdot 10^{-3}) =
5000/100 = 50$. The bandwidth is $\Delta\omega = \omega_{0}/Q = 10^{5}/50 =
2000\,\si{\radian\per\second}$, or $\Delta f \approx 318\,\si{\hertz}$. The
two half-power frequencies sit at approximately $f_{0} \pm \Delta f/2
\approx 15.9\,\si{\kilo\hertz} \pm 159\,\si{\hertz}$. A tank with $Q = 50$
passes a band about $318\,\si{\hertz}$ wide centred at $15.9\,\si{\kilo
\hertz}$ and rejects everything else, the selectivity that lets a radio
front end pick one station out of the crowded spectrum. Raising $R$ (reducing
the tank's loss) raises $Q$ and narrows the band; the loaded $Q$ that a real
receiver sees is lower than the tank's own $Q$ because the following stage
loads the tank with additional resistance.

\subsection{Combining phasors and correcting for a differential measurement}

Section 3.4 added two phasors and section 3.9 read Kirchhoff's voltage law as
a phasor sum. Two further manipulations recur often enough to work
explicitly: summing several phasors of unequal amplitude and phase, and
subtracting one phasor from another to form a differential quantity.

\paragraph{Worked example: the resultant of four unequal phasors.} Four
sinusoidal currents at a common frequency feed one node:
$\tilde I_{1} = 6\angle 0^{\circ}$, $\tilde I_{2} = 4\angle 60^{\circ}$,
$\tilde I_{3} = 3\angle 150^{\circ}$, and $\tilde I_{4} = 2\angle
-90^{\circ}$ amperes, where $A\angle\phi$ abbreviates $A e^{i\phi}$. Their
sum is the current leaving the node. Add in rectangular form, converting each
term:
\begin{align*}
\tilde I_{1} &= 6.000 + 0.000\,i, &
\tilde I_{2} &= 2.000 + 3.464\,i, \\
\tilde I_{3} &= -2.598 + 1.500\,i, &
\tilde I_{4} &= 0.000 - 2.000\,i.
\end{align*}
Summing the real parts gives $6.000 + 2.000 - 2.598 + 0.000 = 5.402$; summing
the imaginary parts gives $0.000 + 3.464 + 1.500 - 2.000 = 2.964$. The
resultant is $\tilde I \approx 5.402 + 2.964\,i$, of magnitude
$\sqrt{5.402^{2} + 2.964^{2}} \approx 6.16\,\si{\ampere}$ and argument
$\arctan(2.964/5.402) \approx 0.502\,\text{rad} \approx 28.8^{\circ}$. The
tip-to-tail polygon of figure~\ref{fig:vol02ch03:phasor-polygon} is the same
sum drawn as a closing vector. The rule of section 3.3 governs the work: add
in rectangular form, because the sum of magnitudes and the sum of angles is
not the magnitude and angle of the sum.

\paragraph{Worked example: line voltage from phase voltages by subtraction.}
A balanced three-phase source has phase voltages (line-to-neutral)
$\tilde V_{a} = 230\angle 0^{\circ}$, $\tilde V_{b} = 230\angle -120^{\circ}$,
and $\tilde V_{c} = 230\angle +120^{\circ}$ volts RMS. The line-to-line
voltage between phases $a$ and $b$ is the phasor difference $\tilde V_{ab} =
\tilde V_{a} - \tilde V_{b}$, not a scalar subtraction of magnitudes. Convert
and subtract:
\[
\tilde V_{a} - \tilde V_{b}
= 230(1 + 0\,i) - 230(-0.5 - 0.866\,i)
= 230(1.5 + 0.866\,i).
\]
The magnitude is $230\sqrt{1.5^{2} + 0.866^{2}} = 230\sqrt{2.25 + 0.75} =
230\sqrt{3} \approx 398\,\si{\volt}$, and the argument is
$\arctan(0.866/1.5) \approx 30^{\circ}$. The line-to-line voltage exceeds the
phase voltage by the factor $\sqrt{3}$, the number every power engineer
carries, and it leads the phase-$a$ voltage by $30^{\circ}$. The factor and
the phase shift are the geometry of subtracting two phasors $120^{\circ}$
apart, and a differential measurement that treated the line voltage as
$230 + 230 = 460\,\si{\volt}$ or as $230 - 230 = 0$ would be wrong by the
same phasor geometry that governs any difference of two sinusoids.

\subsection{Solving a trigonometric equation over a full period}

A trigonometric equation typically has several solutions in one period, and a
calculation that reports only the principal value of an inverse function
misses the rest. The reduction discipline of section 3.1 recovers the full
set.

\paragraph{Worked example: all solutions of a cosine equation.} Find every
$\theta$ in $[0, 2\pi)$ satisfying $2\cos^{2}\theta - \cos\theta - 1 = 0$.
Treat the equation as quadratic in $u = \cos\theta$: $2u^{2} - u - 1 = 0$
factors as $(2u + 1)(u - 1) = 0$, so $u = 1$ or $u = -1/2$. The first,
$\cos\theta = 1$, has the single solution $\theta = 0$ in the period. The
second, $\cos\theta = -1/2$, holds where the reference angle is $\pi/3$ and
the cosine is negative, the second and third quadrants: $\theta = \pi -
\pi/3 = 2\pi/3$ and $\theta = \pi + \pi/3 = 4\pi/3$. The complete solution
set is $\{0,\ 2\pi/3,\ 4\pi/3\}$. A reader who solved $\cos\theta = -1/2$ by
$\arccos(-1/2) = 2\pi/3$ and stopped would report two thirds of the answer,
missing the third-quadrant root that the arccosine's principal range on
$[0, \pi]$ cannot reach. The check is to substitute each candidate back;
each drives the left-hand side to zero.

\subsection{Application: order tracking a run-up on rotating machinery}

A rotating machine radiates vibration at its shaft frequency and at integer
multiples of it, the \engterm{orders}. A turbine spinning at a shaft rate
$f_{s}$ vibrates at $1\times$ (once per revolution, from residual imbalance),
at $2\times$ (from shaft misalignment), and at higher orders from blade
passing, gear meshing, and bearing defects. Diagnosing a fault means
separating these order components, and the trigonometric and phasor
apparatus of this chapter is the tool that does it.

\paragraph{The measurement.} An accelerometer on the bearing housing produces
a voltage proportional to vibration. A once-per-revolution optical tachometer
produces one pulse per shaft turn, which fixes the instantaneous shaft phase.
During a run-up the shaft accelerates from rest to operating speed, so the
shaft frequency $f_{s}(t)$ is not constant and a fixed-frequency phasor does
not apply directly. The order-tracking procedure references every measurement
to the shaft phase rather than to absolute time, which turns the varying
shaft frequency back into a fixed reference in the order domain.

\paragraph{Extracting one order as a phasor.} Consider a single order $m$.
Its contribution to the acceleration is $a_{m}(t) = A_{m}\cos(m\vartheta(t) +
\varphi_{m})$, where $\vartheta(t)$ is the shaft angle read from the
tachometer, integrated pulse by pulse. Multiplying the measured signal by
$e^{-i m\vartheta(t)}$ and averaging over a shaft revolution isolates the
order-$m$ phasor, by exactly the product-to-sum mechanism of section 3.1: the
product $\cos(m\vartheta)\,e^{-i m\vartheta}$ contains a constant term (the
order's own contribution) and terms at $\pm 2m\vartheta$ that the
revolution-average discards. The surviving complex number is the phasor
\[
\tilde A_{m} = \frac{1}{\Theta}\int a(\vartheta)\,e^{-i m\vartheta}\,d\vartheta,
\]
the amplitude and phase of order $m$, with $\Theta = 2\pi$ over one
revolution. This is the Fourier-series coefficient of section 3.4 with shaft
angle in place of time; the order domain is the harmonic domain referenced to
rotation.

\paragraph{Reading the fault.} A machine in good health shows a dominant
$1\times$ order and small higher orders. A rising $1\times$ phasor magnitude
signals developing imbalance; a $1\times$ phasor whose \emph{phase} rotates as
the machine runs up through a shaft critical speed traces the classic
resonance phase shift, and the speed at which the $1\times$ phase passes
through $90^{\circ}$ locates the critical speed. A growing $2\times$ order
signals misalignment; a family of orders clustered around a non-integer
multiple of shaft speed signals a rolling-element bearing defect whose
characteristic frequency is not shaft-locked. The diagnostic reads the
trajectory of each order's phasor as the machine changes speed, exactly the
analytic-signal picture of the section 3.4 mastery box, applied one order at a
time. The balancing correction that follows is itself a phasor calculation:
the trial-weight method adds a known mass at a known angle, measures the
change in the $1\times$ phasor, and solves the linear phasor equation
$\tilde A_{\text{final}} = \tilde A_{\text{original}} + \tilde H\,\tilde W$
for the correction weight $\tilde W$, where the \engterm{influence
coefficient} $\tilde H$ is the complex ratio of vibration change to trial
weight. Field balancing of a turbine rotor is the solution of this one
complex equation, and the entire practice rests on the phasor representation
of a single order of a rotating signal.

\subsection{Application: direction finding by a two-element phase
interferometer}

A pair of antennas separated by a baseline measures the direction to a
distant source by the phase difference between the two received signals. The
apparatus ties together the trigonometry of section 3.1, the phasor of
section 3.4, and the wrap discipline of section 3.7 in one instrument, drawn
in figure~\ref{fig:vol02ch03:phase-interferometer}.

\paragraph{The geometry.} A plane wave from a source at angle $\theta$ from
the array broadside reaches the far antenna after travelling an extra path
$d\sin\theta$, where $d$ is the baseline. The extra path is a fraction
$d\sin\theta/\lambda$ of a wavelength, so the two received signals differ in
phase by
\[
\Delta\varphi = \frac{2\pi}{\lambda}\,d\sin\theta.
\]
Measuring the phase difference and inverting gives the arrival angle
$\theta = \arcsin\!\big(\lambda\,\Delta\varphi/(2\pi d)\big)$. The phase
difference is the measurand; the angle is the inferred quantity. Working the
numbers, take a source at $\lambda = 0.10\,\text{m}$ (a $3\,\si{\giga\hertz}$
signal), a baseline $d = 0.05\,\text{m}$ (half a wavelength), and a measured
$\Delta\varphi = 1.20\,\text{rad}$. Then $\sin\theta = \lambda\,\Delta\varphi/
(2\pi d) = 0.10\cdot 1.20/(2\pi\cdot 0.05) \approx 0.382$, so $\theta \approx
22.5^{\circ}$ from broadside.

\paragraph{The ambiguity.} The phase difference is measured modulo $2\pi$,
the wrap of section 3.7. A baseline of exactly half a wavelength keeps
$\Delta\varphi$ within $\pm\pi$ across the full $\pm 90^{\circ}$ field of
view, so every measured phase maps to one arrival angle and the direction
finding is unambiguous. A longer baseline sharpens the angular resolution,
because a given angular change produces a larger phase change and the phase
is measured to a fixed precision, but it lets $\Delta\varphi$ exceed $\pi$ for
off-broadside sources, so several true angles produce the same wrapped
reading. A baseline of $d = 5\lambda$, ten times the unambiguous value,
resolves angle ten times more finely but wraps ten times across the field of
view.

\paragraph{The coarse-plus-fine resolution.} The standard fix pairs a short
baseline that is unambiguous but coarse with a long baseline that is precise
but wrapped, and uses the coarse estimate to select the correct wrap of the
fine estimate. The coarse baseline reports $\theta$ to within a fraction of
the fine baseline's wrap spacing; the fine baseline then refines it without
ambiguity. This is the same coarse-plus-fine strategy that unwraps a laser
interferometer in section 3.7, where a counted-fringe coarse position selects
the wrap of the fractional-fringe fine phase, and the same strategy that a
multi-baseline radio-astronomy array uses to combine a compact core for wide
unambiguous coverage with long outriggers for high resolution. A single
worked check ties the pieces: with the short $d = \lambda/2$ baseline placing
the source near $22^{\circ}$ and a long $d = 5\lambda$ baseline reporting a
wrapped phase consistent with $22.5^{\circ}$, $58^{\circ}$, or several other
aliases, the coarse reading selects the $22.5^{\circ}$ alias and the fine
baseline pins it to a fraction of a degree. The wrap discipline is not an
afterthought here; it is the design constraint that sets the baseline lengths.

\begin{mastery}
\textbf{From complex rotation to quaternions in three dimensions.} A unit
complex number rotates the plane, and section 3.5 read the rotation as a
$2\times 2$ matrix. Rotations in three dimensions do not commute, and no
single angle captures them, so the two-dimensional complex number does not
extend directly. The object that does is the \engterm{quaternion}, a number
$q = w + x\mathbf{i} + y\mathbf{j} + z\mathbf{k}$ with three imaginary units
satisfying $\mathbf{i}^{2} = \mathbf{j}^{2} = \mathbf{k}^{2} = \mathbf{ijk} =
-1$, from which $\mathbf{ij} = \mathbf{k} = -\mathbf{ji}$ and its cyclic
companions follow. The non-commuting product $\mathbf{ij} \ne \mathbf{ji}$ is
the algebraic image of the fact that rotating about one axis and then another
differs from doing them in the reverse order. A unit quaternion $q$ with
$|q| = 1$ rotates a three-dimensional vector $\mathbf{v}$ (embedded as a pure
quaternion) through the sandwich product $\mathbf{v}' = q\,\mathbf{v}\,q^{-1}$,
by an angle and about an axis read directly off $q = \cos(\vartheta/2) +
\hat{\mathbf{n}}\sin(\vartheta/2)$, the exact three-dimensional analogue of
$e^{i\vartheta} = \cos\vartheta + i\sin\vartheta$ with the half-angle
appearing because the vector is multiplied on both sides. Quaternions are the
working representation of orientation in spacecraft attitude control, robotics,
and computer-graphics animation, because they interpolate smoothly, avoid the
gimbal-lock singularity that the three-angle (Euler-angle) representation
suffers, and need only four numbers and a normalisation rather than a
nine-entry matrix with six constraints. The complex number is the
two-dimensional member of this family, and the quaternion is the step that
buys three-dimensional rotation at the cost of commutativity.
\end{mastery}

\begin{mastery}
\textbf{How the hardware computes a sine: CORDIC.} A processor with no
floating-point multiplier, a pocket calculator, or the arithmetic core of a
software-defined radio does not evaluate $\sin\theta$ from a Taylor series,
because the series needs multiplications the hardware wants to avoid. It uses
\engterm{CORDIC} (coordinate rotation digital computer), an algorithm that
computes a rotation with only additions, subtractions, and bit shifts. The
idea rests on the rotation matrix of section 3.5. A rotation by $\theta$ is
built as a sequence of ever-smaller rotations by fixed angles
$\alpha_{k} = \arctan(2^{-k})$, each of which the algorithm can apply or skip
so that the accumulated angle converges to $\theta$. The key is that
multiplying by $\tan\alpha_{k} = 2^{-k}$ is a bit shift, so each elementary
rotation is
\[
\begin{pmatrix} x_{k+1} \\ y_{k+1} \end{pmatrix}
= \begin{pmatrix} 1 & -s_{k} 2^{-k} \\ s_{k} 2^{-k} & 1 \end{pmatrix}
\begin{pmatrix} x_{k} \\ y_{k} \end{pmatrix},
\]
with the direction $s_{k} = \pm 1$ chosen to drive the running angle toward
$\theta$, and the fixed scale factor $\prod_{k}\cos\alpha_{k} \approx 0.6073$
corrected once at the end. Starting from $(x_{0}, y_{0}) = (K, 0)$ with $K$
the reciprocal scale factor and running the iteration until the residual angle
vanishes leaves $(x, y) = (\cos\theta, \sin\theta)$. Run in the opposite mode,
choosing $s_{k}$ to drive $y_{k}$ to zero instead, the same iteration computes
$\operatorname{atan2}(y_{0}, x_{0})$ and the magnitude $\sqrt{x_{0}^{2} +
y_{0}^{2}}$, so one small piece of logic evaluates both the forward and the
inverse of the rectangular-to-polar conversion of section 3.3. CORDIC is why a
low-cost microcontroller can demodulate a phasor in real time; it is the
rotation matrix of this chapter reduced to shifts and adds.
\end{mastery}

\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-oblique-triangle.tex}
\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-fifth-roots.tex}
\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-parallel-rlc-resonance.tex}
\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-phasor-polygon.tex}
\input{volumes/02-form/ch03-trigonometry-complex-phasors/figures/fig-phase-interferometer.tex}

\subsection{Constructible angles and the fifth roots of unity}

The roots of unity of section 3.3 sit at the corners of a regular polygon.
For a few values of $n$ the corner coordinates are constructible with
straightedge and compass, which is the classical question of which regular
polygons can be drawn exactly. The pentagon is the smallest nontrivial case,
and its fifth roots of unity carry an exact cosine that the reader can derive
from the roots-of-unity algebra alone.
Figure~\ref{fig:vol02ch03:fifth-roots} draws the five roots.

\paragraph{Worked example: the exact value of $\cos(2\pi/5)$.} The five
fifth-roots of unity are $\zeta_{k} = e^{i 2\pi k/5}$ for $k = 0,\dots,4$, and
they satisfy $\sum_{k=0}^{4}\zeta_{k} = 0$ because they are the roots of
$z^{5} - 1 = (z - 1)(z^{4} + z^{3} + z^{2} + z + 1)$ and the coefficient sum
forces the root sum to zero. Setting aside the root $\zeta_{0} = 1$, the other
four satisfy $z^{4} + z^{3} + z^{2} + z + 1 = 0$. Divide by $z^{2}$ and group,
using $z + 1/z = 2\cos(2\pi k/5)$ for a root on the unit circle:
\[
\left(z^{2} + \frac{1}{z^{2}}\right) + \left(z + \frac{1}{z}\right) + 1 = 0.
\]
Write $t = z + 1/z = 2\cos(2\pi k/5)$; then $z^{2} + 1/z^{2} = t^{2} - 2$, and
the equation becomes $t^{2} + t - 1 = 0$, with roots $t = (-1 \pm
\sqrt{5})/2$. The positive root corresponds to $k = 1$ (the smaller angle,
larger cosine), giving
\[
2\cos\frac{2\pi}{5} = \frac{-1 + \sqrt{5}}{2},
\qquad
\cos\frac{2\pi}{5} = \frac{\sqrt{5} - 1}{4} \approx 0.309.
\]
The cosine of the pentagon's central angle is an exact expression in
$\sqrt{5}$, which is the algebraic reason the regular pentagon is
constructible: every coordinate the construction needs lies in a field
reached from the rationals by square roots alone. The negative root of the
quadratic gives $2\cos(4\pi/5) = (-1 - \sqrt{5})/2$, the cosine of the next
vertex. The roots-of-unity algebra has turned a geometric constructibility
question into the solution of one quadratic, and the golden ratio
$(1 + \sqrt{5})/2$ that governs the pentagon appears as the reciprocal of
this cosine's relatives.

\section{Project}\pathtag{core}

\begin{project}[Series RLC steady-state response by hand]
The chapter project, per the dossier and the editorial decision
Q55-vol02-ch03, asks the reader to derive by hand the AC
steady-state response of a series RLC circuit driven by a
sinusoidal source.

\textbf{Track}. Analysis with optional simulation. Hazard class:
none. The analysis is done with paper, pencil, and a calculator;
the simulation, if undertaken, is in a programming environment
(Python, MATLAB, Julia, or equivalent).

\textbf{Tools}. Paper, pencil, calculator. Optionally a numerical
ODE solver for the simulation track.

\textbf{Procedure (analysis track)}.

\begin{enumerate}[leftmargin=*]
  \item \emph{Choose parameters}. Pick values of $R$, $L$, $C$,
    $V_{m}$, and $\omega$ that make the calculation tractable but
    not trivial. A reasonable set is $R = 100 \,\si{\ohm}$, $L =
    50 \,\si{\milli\henry}$, $C = 22 \,\si{\micro\farad}$,
    $V_{m} = 10 \,\si{\volt}$, $\omega = 2\pi \cdot 100 \,\si
    {\hertz}$, but the reader is encouraged to choose
    parameters whose resonant frequency $\omega_{0} = 1/\sqrt{LC}$
    falls within a factor of two of the chosen drive frequency
    $\omega$.

  \item \emph{Derive the impedance}. Compute $Z_{R}$, $Z_{L}$,
    $Z_{C}$, and the series $Z = Z_{R} + Z_{L} + Z_{C}$. Express
    $Z$ in both rectangular form $Z = R + iX$ (where $X = \omega
    L - 1/(\omega C)$ is the \engterm{reactance}) and polar form
    $Z = |Z| e^{i\varphi_{Z}}$.

  \item \emph{Derive the current phasor}. Compute $\tilde{I} =
    \tilde{V}_{s}/Z$ in polar form. State the current's amplitude
    $|\tilde{I}|$ and its phase angle relative to the source.

  \item \emph{Derive each element voltage}. Compute
    $\tilde{V}_{R}$, $\tilde{V}_{L}$, and $\tilde{V}_{C}$ from the
    current phasor and the corresponding impedance. State each
    voltage's amplitude and phase angle relative to the source.

  \item \emph{Verify Kirchhoff}. Compute $\tilde{V}_{R} +
    \tilde{V}_{L} + \tilde{V}_{C}$ and verify it equals
    $\tilde{V}_{s}$ within rounding error.

  \item \emph{Compute the average power}. Compute $P =
    \tfrac{1}{2}\Re[\tilde{V}_{s}\bar{\tilde{I}}]$ and verify it
    equals $|\tilde{I}|^{2} R / 2$, the power dissipated in the
    resistor.

  \item \emph{Convert each phasor back}. Write $i(t)$, $v_{R}(t)$,
    $v_{L}(t)$, and $v_{C}(t)$ as time-domain sinusoids.
\end{enumerate}

\textbf{Procedure (optional simulation track)}.

\begin{enumerate}[leftmargin=*]
  \item Set up the second-order ODE for the series RLC circuit:
    $L\,d^{2}q/dt^{2} + R\,dq/dt + q/C = V_{m}\cos(\omega t)$,
    where $q$ is the capacitor charge. (The reader who has not
    yet covered ODEs may convert by writing $i = dq/dt$ and
    $v_{C} = q/C$, yielding a coupled first-order system in
    $(i, q)$.)
  \item Integrate numerically using a standard ODE solver. Run
    long enough that the transient response (the response in
    the first few cycles after $t = 0$, set by the homogeneous
    solution) has died out and the steady-state response
    dominates.
  \item Compare the steady-state numerical solution to the
    phasor-derived solution. Plot both on the same axes for
    one or two cycles after the transient has settled. Compute
    the maximum discrepancy and identify its source (numerical
    error, transient residual, parameter mismatch).
\end{enumerate}

\textbf{Reflection (1500 words)}. The reader writes a 1500-word
narrative addressing:

\begin{itemize}[leftmargin=*]
  \item What the phasor representation buys over time-domain
    integration: how the analysis track reduced a second-order
    ODE to a complex algebraic equation, and what would be lost
    in a direct time-domain solution that did not invoke
    phasors.
  \item Where the phasor representation does not apply: the
    transient response, multi-frequency drives, nonlinear
    elements, time-varying parameters.
  \item How the resonance condition $\omega L = 1/(\omega C)$
    appears in the phasor calculation, and what happens to the
    impedance, current, and element voltages at resonance.
  \item Whether the series RLC circuit is a stable system, and
    in what sense the archetype of stability (rotation as
    preserved structure, magnitude as conserved quantity over a
    cycle) applies to it.
  \item One real engineering system the reader has encountered
    or can imagine in which the phasor representation would be
    the working tool of the analysis, and one in which it would
    not.
\end{itemize}

\textbf{Deliverable}. The hand-derived solution, the algebraic
verification of Kirchhoff's law and of the power balance, the
time-domain expressions for current and each element voltage,
and the 1500-word reflection. The optional simulation deliverable
includes the source code, the parameter set, the time-domain
plots, and the comparison to the phasor result.

\textbf{Time}. Analysis: three to four hours. Optional simulation:
two to three hours. Reflection: two to three hours.

\textbf{Phases.} The work proceeds in five named phases. Phase
boundaries are the place to stop, check, and decide whether to
proceed.

\begin{enumerate}[leftmargin=*]
  \item \emph{Phase 1 (parameter check, 20 minutes)}. Choose
    parameters and compute the resonant frequency $\omega_{0}$, the
    drive frequency $\omega$, and the ratio $\omega/\omega_{0}$.
    Verify the chosen drive frequency is within a factor of two of
    resonance (so the calculation exercises both the reactive
    cancellation and the resistive dissipation, rather than running
    in a purely reactive or purely resistive limit). Compute the
    quality factor $Q = \omega_{0} L / R$ as a single number to
    track through the rest of the calculation.

  \item \emph{Phase 2 (impedance and current, 45 minutes)}. Compute
    $Z_{R}$, $Z_{L}$, $Z_{C}$, the series $Z$, and the current
    phasor $\tilde I = \tilde V_{s} / Z$. Express each quantity in
    both rectangular and polar form. Sketch the impedance triangle
    in the complex plane.

  \item \emph{Phase 3 (element voltages and KVL, 45 minutes)}.
    Compute $\tilde V_{R}, \tilde V_{L}, \tilde V_{C}$ from the
    current phasor and the corresponding impedance. Sketch the
    phasor diagram (the four voltage phasors and the closure to
    $\tilde V_{s}$) at a scale that makes the magnitudes legible.
    Verify $\tilde V_{R} + \tilde V_{L} + \tilde V_{C} = \tilde V_{s}$
    within rounding error.

  \item \emph{Phase 4 (power, 30 minutes)}. Compute the average
    power $P = \tfrac{1}{2}\Re[\tilde V_{s} \bar{\tilde I}]$ and
    verify it equals $|\tilde I|^{2} R / 2$. Compute the reactive
    power $Q = \tfrac{1}{2}\Im[\tilde V_{s} \bar{\tilde I}]$ and
    verify it equals the algebraic sum of the inductive and
    capacitive contributions. Compute the apparent power $|S|$ and
    the power factor $P / |S|$.

  \item \emph{Phase 5 (time-domain reconstruction and reflection,
    one to two hours)}. Write $i(t)$, $v_{R}(t)$, $v_{L}(t)$,
    $v_{C}(t)$ as time-domain sinusoids and verify by direct
    substitution that the KVL equation $v_{R}(t) + v_{L}(t) +
    v_{C}(t) = v_{s}(t)$ holds at $t = 0$ and at $t = T/4$. Write
    the 1500-word reflection.
\end{enumerate}

\textbf{Success criteria.} A deliverable meets the project bar if
the following are all true.

\begin{itemize}[leftmargin=*]
  \item The chosen parameters carry units throughout and the
    impedance, current, voltage, and power values are all in SI
    units with the appropriate prefix (ohm, ampere, volt, watt).
  \item The phasor diagram is drawn to scale and visibly closes:
    the chain $\tilde V_{R} + \tilde V_{L} + \tilde V_{C}$ ends on
    $\tilde V_{s}$ within the drawing's line width.
  \item The numerical KVL check $\tilde V_{R} + \tilde V_{L} +
    \tilde V_{C} = \tilde V_{s}$ holds to at least three significant
    figures.
  \item The two power expressions $P = \tfrac{1}{2}\Re[\tilde V_{s}
    \bar{\tilde I}]$ and $P = |\tilde I|^{2} R / 2$ agree to at
    least three significant figures.
  \item The reflection identifies one concrete LTI system class for
    which phasor analysis is the working tool, names the resonance
    condition $\omega L = 1/(\omega C)$ as a single line in the
    impedance expression, and explains where the representation
    breaks (transients, nonlinear elements, multi-frequency drive).
  \item For the optional simulation track: the steady-state portion
    of the numerical current trace agrees with the phasor-derived
    current to within $1\%$ relative error after the transient has
    decayed.
\end{itemize}

\textbf{Common failure modes.} The deliverables that miss the bar
typically fail in one of the following ways. Each failure mode is
listed with the diagnosis and the recovery.

\begin{itemize}[leftmargin=*]
  \item \emph{Sign error on the capacitive reactance.} The
    capacitor impedance is $Z_{C} = -i / (\omega C)$, not
    $+i / (\omega C)$. The error puts the capacitor voltage on the
    wrong side of the real axis and breaks KVL by twice the
    capacitor voltage magnitude. Recovery: re-derive $Z_{C}$ from
    $i_{C} = C\,d v_{C}/dt$ and the phasor differentiation rule;
    confirm the sign by sketching the capacitor's voltage and
    current as a function of time and noting that the current
    leads.

  \item \emph{Confusion between current phase and impedance phase.}
    The current phasor's argument is the negative of the
    impedance's argument: $\tilde I = \tilde V_{s}/Z$, so
    $\arg(\tilde I) = \arg(\tilde V_{s}) - \arg(Z)$. A deliverable
    that reports the current's phase as $+\varphi_{Z}$ rather than
    $-\varphi_{Z}$ has confused the two. Recovery: write the
    division explicitly in polar form.

  \item \emph{Off-by-one-half in the average-power formula.} The
    peak-value convention gives $P = \tfrac{1}{2}\Re[\tilde V
    \bar{\tilde I}]$; the RMS-value convention gives
    $P = \Re[\tilde V \bar{\tilde I}]$ with no factor. A deliverable
    that uses the peak convention but forgets the factor of one
    half overstates the dissipated power by a factor of two.
    Recovery: cross-check with $|\tilde I|^{2} R / 2$ (or
    $|\tilde I_{\text{RMS}}|^{2} R$); the answer should agree.

  \item \emph{Quadrant error in $\varphi_{Z}$.} The
    $\arctan$ function returns a value in $(-\pi/2, \pi/2)$. A
    reactance computation that produces a positive imaginary part
    needs the angle in the first quadrant; one that produces a
    negative imaginary part needs the angle in the fourth quadrant.
    A reader who writes $\varphi_{Z} = \arctan(X/R)$ without
    inspecting the sign of $X$ gets the right answer for $X > 0$ by
    accident and the wrong answer for $X < 0$ that is consistent
    with $\arctan$ returning a fourth-quadrant value, which agrees
    with $\operatorname{atan2}(X, R)$ here because $R > 0$. The
    confusion becomes load-bearing when $R \to 0$. Recovery: use
    $\operatorname{atan2}(X, R)$ throughout.

  \item \emph{Misread of $\omega$ versus $f$.} The drive frequency
    $\omega$ is in radians per second; the frequency $f$ is in
    hertz; $\omega = 2\pi f$. A reader who substitutes $f$
    where $\omega$ belongs reports an impedance and current that
    are off by a factor of $2\pi$. Recovery: write the conversion
    line $\omega = 2\pi f$ explicitly at the start of phase 1 and
    track units through every step.
\end{itemize}
\end{project}

\section{Exercises}

The exercises move from trigonometric arithmetic and identity
manipulation through complex-number algebra, polar conversions,
and phasor calculations to design and judgement on the project's
working machinery. Closed problems have full solutions; design
and judgement problems have rubrics or reference write-ups in the
companion notes.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Convert each angle to the other unit. Express in radians as a
fraction of $\pi$ where possible and as a decimal otherwise: (a)
$30^{\circ}$; (b) $135^{\circ}$; (c) $-45^{\circ}$; (d) $720^{\circ}$;
(e) $1 \,\text{rad}$.

\paragraph{Solution sketch.}
Multiply by $\pi/180$ to go from degrees to radians.
(a) $30^{\circ} = \pi/6 \approx 0.5236\,\text{rad}$.
(b) $135^{\circ} = 3\pi/4 \approx 2.356\,\text{rad}$.
(c) $-45^{\circ} = -\pi/4 \approx -0.7854\,\text{rad}$.
(d) $720^{\circ} = 4\pi \approx 12.566\,\text{rad}$ (two full
revolutions; the same point on the unit circle as $0$).
(e) $1\,\text{rad} \cdot (180/\pi) \approx 57.296^{\circ}$, or
about $57^{\circ}17'45''$. The radian is the natural unit for
analytic formulas because the Taylor series for sine and cosine
($\sin\theta = \theta - \theta^{3}/6 + \cdots$) have unit slope
at the origin only when the argument is in radians.
\end{exercise}

\begin{exercise}
Compute, exactly when possible: (a) $\sin(7\pi/6)$; (b) $\cos(5\pi/4)$;
(c) $\tan(2\pi/3)$; (d) $\sin(11\pi/6) + \cos(11\pi/6)$. State the
quadrant of each angle on the unit circle.

\paragraph{Solution sketch.}
(a) $7\pi/6$ is in the third quadrant; $\sin$ is negative there.
$\sin(7\pi/6) = -\sin(\pi/6) = -1/2$.
(b) $5\pi/4$ is in the third quadrant; $\cos$ is negative.
$\cos(5\pi/4) = -\cos(\pi/4) = -\sqrt{2}/2$.
(c) $2\pi/3$ is in the second quadrant; $\sin$ positive, $\cos$
negative, so $\tan$ negative. $\tan(2\pi/3) = \sin(2\pi/3)/\cos(2\pi/3)
= (\sqrt{3}/2)/(-1/2) = -\sqrt{3}$.
(d) $11\pi/6$ is in the fourth quadrant. $\sin(11\pi/6) = -1/2$,
$\cos(11\pi/6) = \sqrt{3}/2$, so the sum is $(\sqrt{3} - 1)/2
\approx 0.366$.
\end{exercise}

\begin{exercise}
A right triangle has legs of length $5$ and $12$. Compute the
hypotenuse, the sine and cosine of the angle opposite the leg of
length $5$, and the angle in degrees.

\paragraph{Solution sketch.}
By Pythagoras, the hypotenuse is $\sqrt{5^{2} + 12^{2}} = \sqrt{169}
= 13$. The angle $\theta$ opposite the leg of length $5$ satisfies
$\sin\theta = 5/13 \approx 0.3846$ and $\cos\theta = 12/13 \approx
0.9231$. Then $\theta = \arcsin(5/13) \approx 0.3948\,\text{rad}
\approx 22.62^{\circ}$. The triangle $5$-$12$-$13$ is the second
canonical Pythagorean triple after $3$-$4$-$5$; the reader who
notices the triple finishes the exercise in three lines.
\end{exercise}

\begin{exercise}
Express each complex number in both rectangular form $a + bi$
and polar form $r e^{i\theta}$ with $\theta \in (-\pi, \pi]$:
(a) $z = -3 + 3i$; (b) $z = 2 e^{i\pi/3}$; (c) $z = (1 + i)^{4}$;
(d) $z = 1/(1 - i)$.

\paragraph{Solution sketch.}
(a) Rectangular is given. $r = \sqrt{9 + 9} = 3\sqrt{2}$,
$\theta = \operatorname{atan2}(3, -3) = 3\pi/4$, so polar is
$3\sqrt{2}\,e^{i 3\pi/4}$.
(b) Polar is given. Rectangular: $2(\cos(\pi/3) + i\sin(\pi/3)) =
2(1/2 + i\sqrt{3}/2) = 1 + i\sqrt{3}$.
(c) $(1 + i) = \sqrt{2}\,e^{i\pi/4}$, so $(1 + i)^{4} = (\sqrt{2})^{4}
e^{i\pi} = 4 e^{i\pi} = -4$. Rectangular: $-4 + 0\,i$. Polar:
$4\,e^{i\pi}$.
(d) Multiply by the conjugate: $1/(1 - i) = (1 + i)/((1 - i)(1 + i))
= (1 + i)/2 = 1/2 + (1/2)i$. Polar: $r = \sqrt{1/2}/1 = (\sqrt{2}/2)$,
$\theta = \pi/4$, so $(\sqrt{2}/2)\,e^{i\pi/4}$.
\end{exercise}

\begin{exercise}
Compute, expressing each result in rectangular form: (a) $(3 +
4i)(1 - 2i)$; (b) $(2 - i)/(1 + 3i)$; (c) $(1 + i)^{6}$; (d)
$\overline{(1 + 2i)(3 - i)}$.

\paragraph{Solution sketch.}
(a) $(3 + 4i)(1 - 2i) = 3 - 6i + 4i - 8i^{2} = 3 - 2i + 8 =
11 - 2i$.
(b) Multiply by $\overline{(1 + 3i)} = 1 - 3i$ over itself:
$(2 - i)(1 - 3i) = 2 - 6i - i + 3i^{2} = -1 - 7i$;
$|1 + 3i|^{2} = 1 + 9 = 10$; the quotient is $(-1 - 7i)/10 =
-0.1 - 0.7\,i$.
(c) Use polar: $(1 + i) = \sqrt{2}\,e^{i\pi/4}$, so
$(1 + i)^{6} = (\sqrt{2})^{6}\,e^{i 3\pi/2} = 8\,e^{i 3\pi/2} =
0 - 8i = -8i$.
(d) $(1 + 2i)(3 - i) = 3 - i + 6i - 2i^{2} = 5 + 5i$; the conjugate
is $5 - 5i$.
\end{exercise}

\begin{exercise}
Compute the four fourth roots of $z = -16$. State the magnitude
and the argument of each. Sketch their positions on the unit
circle (scaled by the common magnitude).

\paragraph{Solution sketch.}
Write $z = 16\,e^{i\pi}$. The fourth roots have magnitude
$16^{1/4} = 2$ and arguments $(\pi + 2\pi k)/4$ for $k = 0, 1, 2, 3$:
$\theta \in \{\pi/4, 3\pi/4, 5\pi/4, 7\pi/4\}$ (equivalently,
$\{45^{\circ}, 135^{\circ}, 225^{\circ}, 315^{\circ}\}$ or
$\{\pi/4, 3\pi/4, -3\pi/4, -\pi/4\}$ in principal range). The four
roots are $2\,e^{i\pi/4} = \sqrt{2}(1 + i)$, $2\,e^{i 3\pi/4} =
\sqrt{2}(-1 + i)$, $2\,e^{i 5\pi/4} = \sqrt{2}(-1 - i)$, and
$2\,e^{i 7\pi/4} = \sqrt{2}(1 - i)$. They sit at the corners of
a square of side $2\sqrt{2}$ inscribed in a circle of radius
$2$. Multiplying any of the four roots by itself four times
returns $-16$ as a check.
\end{exercise}

\begin{exercise}
A sinusoidal voltage is $v(t) = 12\cos(2\pi \cdot 60 \,t -
\pi/4) \,\si{\volt}$. State the peak amplitude, the angular
frequency in radians per second, the frequency in hertz, the
period in milliseconds, the initial phase in radians and in
degrees, and the phasor in polar and rectangular form.

\paragraph{Solution sketch.}
Peak amplitude $V_{m} = 12\,\si{\volt}$. Angular frequency
$\omega = 2\pi \cdot 60 \approx 377\,\si{\radian\per\second}$.
Frequency $f = 60\,\si{\hertz}$. Period $T = 1/f \approx
16.67\,\si{\milli\second}$. Initial phase $\varphi = -\pi/4
\approx -0.7854\,\text{rad} \approx -45^{\circ}$. Phasor (peak
convention): polar form $\tilde V = 12\,e^{-i\pi/4}\,\si{\volt}$;
rectangular form $\tilde V = 12\cos(-\pi/4) + i \cdot 12\sin(-\pi/4)
= 12 (\sqrt{2}/2) - i \cdot 12 (\sqrt{2}/2) = 6\sqrt{2} -
i\,6\sqrt{2} \approx 8.485 - 8.485\,i \,\si{\volt}$.
\end{exercise}

\begin{exercise}
Add two sinusoids at the same frequency by phasor: $x_{1}(t) =
5\cos(\omega t)$ and $x_{2}(t) = 12\cos(\omega t + \pi/2)$.
Express the sum as a single sinusoid $A\cos(\omega t + \varphi)$,
giving $A$ and $\varphi$ to four significant figures.

\paragraph{Solution sketch.}
Phasors: $\tilde X_{1} = 5$, $\tilde X_{2} = 12\,e^{i\pi/2} =
12i$. Sum: $\tilde X = 5 + 12i$. Magnitude: $|\tilde X| =
\sqrt{25 + 144} = \sqrt{169} = 13.00$ (the Pythagorean triple
$5$-$12$-$13$ again). Argument: $\varphi = \arctan(12/5) \approx
1.176\,\text{rad} \approx 67.38^{\circ}$. The sum is
$x_{1}(t) + x_{2}(t) = 13.00\cos(\omega t + 1.176)$.
\end{exercise}

\begin{exercise}
A series RL circuit has $R = 50 \,\si{\ohm}$ and $L = 20 \,\si
{\milli\henry}$. Compute the impedance at $\omega = 2\pi \cdot 1
\,\si{\kilo\hertz}$ in both rectangular and polar form. State
the phase angle by which the current lags the voltage.

\paragraph{Solution sketch.}
$\omega = 2\pi \cdot 1000 \approx 6283\,\si{\radian\per\second}$;
$\omega L = 6283 \cdot 0.020 \approx 125.7\,\si{\ohm}$.
Rectangular: $Z = R + i\omega L = 50 + 125.7\,i\,\si{\ohm}$. Polar:
$|Z| = \sqrt{50^{2} + 125.7^{2}} \approx 135.3\,\si{\ohm}$,
$\arg Z = \arctan(125.7/50) \approx 1.193\,\text{rad} \approx
68.30^{\circ}$. The current lags the voltage by the impedance's
phase angle, $\approx 68.3^{\circ}$. The inductor dominates the
resistor at this frequency; the circuit is close to purely
inductive.
\end{exercise}

\begin{exercise}
A capacitor of $C = 1 \,\si{\micro\farad}$ is driven by a sinusoidal
voltage of amplitude $10 \,\si{\volt}$ at $\omega = 2\pi \cdot 50
\,\si{\hertz}$. Compute the impedance, the current amplitude, and
the phase angle by which the current leads the voltage. Compute
the average power dissipated in the capacitor.

\paragraph{Solution sketch.}
$\omega = 2\pi \cdot 50 \approx 314.2\,\si{\radian\per\second}$;
$1/(\omega C) = 1/(314.2 \cdot 10^{-6}) \approx 3183\,\si{\ohm}$.
Impedance $Z_{C} = -i/(\omega C) \approx -3183\,i\,\si{\ohm}$, polar
form $3183\,e^{-i\pi/2}\,\si{\ohm}$. Current amplitude
$|\tilde I| = |\tilde V|/|Z_{C}| = 10/3183 \approx 3.14\,\si
{\milli\ampere}$. The current leads the voltage by $\pi/2 =
90^{\circ}$ (current phase is voltage phase minus impedance phase,
$0 - (-\pi/2) = +\pi/2$). The average power dissipated in an
ideal capacitor is zero: $P = \tfrac{1}{2}\Re[\tilde V \bar{\tilde
I}] = \tfrac{1}{2}\Re[10 \cdot (-i \cdot 3.14 \times 10^{-3})] =
\tfrac{1}{2}\Re[-31.4 \times 10^{-3}\,i] = 0$. The capacitor
exchanges energy with the source but dissipates none; the
energy flow is reactive, $|Q| = (1/2)|\tilde V||\tilde I| \approx
15.7\,\si{\milli\text{var}}$.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Starting from the angle-sum identity for cosine, derive the
angle-difference identity for cosine and the double-angle
identity for cosine. Then derive the half-angle identity
$\cos^{2}\theta = (1 + \cos 2\theta)/2$.

\paragraph{Solution sketch.}
Angle-sum: $\cos(\alpha + \beta) = \cos\alpha\cos\beta -
\sin\alpha\sin\beta$. Substitute $\beta \to -\beta$ and use
$\cos(-\beta) = \cos\beta$, $\sin(-\beta) = -\sin\beta$:
$\cos(\alpha - \beta) = \cos\alpha\cos\beta + \sin\alpha\sin\beta$.
Set $\alpha = \beta = \theta$ in the angle-sum identity:
$\cos(2\theta) = \cos^{2}\theta - \sin^{2}\theta$. Combine with
$\sin^{2}\theta = 1 - \cos^{2}\theta$ (Pythagorean): $\cos(2\theta)
= 2\cos^{2}\theta - 1$, so $\cos^{2}\theta = (1 + \cos 2\theta)/2$.
\end{exercise}

\begin{exercise}
Using Euler's identity $e^{i\theta} = \cos\theta + i\sin\theta$,
prove the angle-sum identities for cosine and sine by expanding
$e^{i(\alpha + \beta)} = e^{i\alpha} e^{i\beta}$ and equating real
and imaginary parts.

\paragraph{Solution sketch.}
$e^{i(\alpha + \beta)} = \cos(\alpha + \beta) + i\sin(\alpha +
\beta)$ by Euler. On the right, $e^{i\alpha} e^{i\beta} =
(\cos\alpha + i\sin\alpha)(\cos\beta + i\sin\beta) =
\cos\alpha\cos\beta - \sin\alpha\sin\beta + i(\sin\alpha\cos\beta
+ \cos\alpha\sin\beta)$. Equating real parts: $\cos(\alpha + \beta)
= \cos\alpha\cos\beta - \sin\alpha\sin\beta$. Equating imaginary
parts: $\sin(\alpha + \beta) = \sin\alpha\cos\beta + \cos\alpha
\sin\beta$. The angle-sum identities are the real and imaginary
parts of one line of complex multiplication.
\end{exercise}

\begin{exercise}
Using the Euler formulas $\cos\theta = (e^{i\theta} +
e^{-i\theta})/2$ and $\sin\theta = (e^{i\theta} - e^{-i\theta})/(2i)$,
prove the product-to-sum identity $\cos\alpha\cos\beta =
\tfrac{1}{2}[\cos(\alpha - \beta) + \cos(\alpha + \beta)]$ by
direct multiplication.

\paragraph{Solution sketch.}
$\cos\alpha\cos\beta = \tfrac{1}{4}(e^{i\alpha} + e^{-i\alpha})
(e^{i\beta} + e^{-i\beta}) = \tfrac{1}{4}(e^{i(\alpha + \beta)} +
e^{i(\alpha - \beta)} + e^{-i(\alpha - \beta)} + e^{-i(\alpha +
\beta)})$. The first and fourth terms combine to $2\cos(\alpha
+ \beta)$; the second and third to $2\cos(\alpha - \beta)$. The
sum is $\tfrac{1}{4}[2\cos(\alpha + \beta) + 2\cos(\alpha -
\beta)] = \tfrac{1}{2}[\cos(\alpha - \beta) + \cos(\alpha +
\beta)]$.
\end{exercise}

\begin{exercise}
Show that $|z_{1} z_{2}| = |z_{1}|\,|z_{2}|$ for any two complex
numbers $z_{1}, z_{2}$. Hint: write $z_{1} z_{2} \overline{(z_{1}
z_{2})}$ and use the fact that conjugation distributes over
multiplication.

\paragraph{Solution sketch.}
$|z_{1} z_{2}|^{2} = (z_{1} z_{2})\overline{(z_{1} z_{2})} =
(z_{1} z_{2})(\bar z_{1} \bar z_{2}) = (z_{1} \bar z_{1})(z_{2}
\bar z_{2}) = |z_{1}|^{2} |z_{2}|^{2}$. Taking the positive square
root: $|z_{1} z_{2}| = |z_{1}||z_{2}|$. The fact uses
$\overline{wz} = \bar w \bar z$, which the reader verifies in one
line by writing $w = a + bi$, $z = c + di$ and expanding both sides.
\end{exercise}

\begin{exercise}
Show that the rotation matrix $R(\theta) = \begin{psmallmatrix}
\cos\theta & -\sin\theta \\ \sin\theta & \cos\theta
\end{psmallmatrix}$ satisfies $R(\alpha) R(\beta) = R(\alpha +
\beta)$ by direct multiplication. State which trigonometric
identities the calculation reproduces.

\paragraph{Solution sketch.}
The $(1, 1)$ entry of $R(\alpha) R(\beta)$ is $\cos\alpha\cos\beta
+ (-\sin\alpha)(\sin\beta) = \cos\alpha\cos\beta - \sin\alpha
\sin\beta$, which equals $\cos(\alpha + \beta)$ by the cosine
angle-sum identity. The $(2, 1)$ entry is $\sin\alpha\cos\beta +
\cos\alpha\sin\beta = \sin(\alpha + \beta)$ by the sine angle-sum
identity. The $(1, 2)$ and $(2, 2)$ entries reduce to
$-\sin(\alpha + \beta)$ and $\cos(\alpha + \beta)$ by the same
identities with $\beta \to -\beta$. The matrix identity is the
angle-sum identities packaged into a single equation in $2\times
2$ matrices.
\end{exercise}

\begin{exercise}
Derive the impedance of an inductor $Z_{L} = i\omega L$ from
the time-domain rule $v_{L}(t) = L\,di_{L}/dt$ by writing
$i_{L}(t) = \Re[\tilde{I}_{L} e^{i\omega t}]$ and computing
$v_{L}(t)$. Identify which step uses the linearity of the real
part operator and which uses the differentiation rule.

\paragraph{Solution sketch.}
$v_{L}(t) = L\,d/dt\,\Re[\tilde I_{L} e^{i\omega t}] = L\,\Re[d/dt
(\tilde I_{L} e^{i\omega t})]$. The interchange of $\Re$ and
$d/dt$ uses the linearity of the real-part operator. Then
$d/dt(\tilde I_{L} e^{i\omega t}) = i\omega \tilde I_{L} e^{i\omega
t}$ uses the differentiation rule on the complex exponential. So
$v_{L}(t) = L \cdot \Re[i\omega \tilde I_{L} e^{i\omega t}] =
\Re[i\omega L \tilde I_{L} e^{i\omega t}]$. Reading off the
phasor: $\tilde V_{L} = i\omega L \tilde I_{L}$, hence $Z_{L} =
\tilde V_{L} / \tilde I_{L} = i\omega L$.
\end{exercise}

\begin{exercise}
Derive the formula $|\tilde{Y}_{1} + \tilde{Y}_{2}| = \sqrt{A_{1}^{2}
+ A_{2}^{2} + 2 A_{1} A_{2}\cos(\varphi_{2} - \varphi_{1})}$
for the magnitude of a sum of two phasors, by expanding
$|\tilde{Y}_{1} + \tilde{Y}_{2}|^{2} = (\tilde{Y}_{1} +
\tilde{Y}_{2})\overline{(\tilde{Y}_{1} + \tilde{Y}_{2})}$.

\paragraph{Solution sketch.}
Write $\tilde Y_{k} = A_{k} e^{i\varphi_{k}}$. Expand:
$|\tilde Y_{1} + \tilde Y_{2}|^{2} = (\tilde Y_{1} + \tilde
Y_{2})(\bar{\tilde Y}_{1} + \bar{\tilde Y}_{2}) = |\tilde Y_{1}|^{2}
+ |\tilde Y_{2}|^{2} + \tilde Y_{1} \bar{\tilde Y}_{2} +
\bar{\tilde Y}_{1} \tilde Y_{2}$. The first two terms are $A_{1}^{2}
+ A_{2}^{2}$. The cross terms are $A_{1} A_{2} (e^{i(\varphi_{1}
- \varphi_{2})} + e^{-i(\varphi_{1} - \varphi_{2})}) = 2 A_{1}
A_{2} \cos(\varphi_{1} - \varphi_{2}) = 2 A_{1} A_{2} \cos(\varphi_{2}
- \varphi_{1})$ (cosine is even). Taking the positive square root
gives the stated formula, the law of cosines applied to the
phasor triangle.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the magnitude and the argument of $z = 7 - 24i$ by
inspection, before computing. State your estimate, then compute
the exact values, and report the relative error of the estimate.

\paragraph{Solution sketch.}
The Pythagorean triple $7$-$24$-$25$ is the secondary-school
standby (one of the small primitive triples); an estimator who
spots it reports magnitude exactly $25$ in one line. An estimator
who does not spot it observes $|z|^{2} = 49 + 576 = 625$ and
extracts $|z| = 25$. The argument is in the fourth quadrant
because the real part is positive and the imaginary part is
negative; $\arctan(-24/7) \approx -1.287\,\text{rad} \approx
-73.74^{\circ}$. The estimator who anchors on the $24/7 \approx
3.43$ slope and recognises this as a steep angle close to but
less than $\pi/2$ would estimate near $-75^{\circ}$; the relative
error is about $2\%$ on the angle and zero on the magnitude.
\end{exercise}

\begin{exercise}
Estimate the cube roots of $z = 8i$ to within a tenth of a
radian by sketching them on the complex plane. Then compute
them exactly and compare.

\paragraph{Solution sketch.}
$z = 8\,e^{i\pi/2}$. The three cube roots have magnitude
$8^{1/3} = 2$ and arguments $(\pi/2 + 2\pi k)/3$ for $k = 0, 1, 2$:
$\theta \in \{\pi/6, 5\pi/6, 9\pi/6\} = \{\pi/6, 5\pi/6, -\pi/2\}$,
spaced by $2\pi/3$ around the circle of radius $2$. In rectangular
form: $w_{0} = 2\,e^{i\pi/6} = \sqrt{3} + i$,
$w_{1} = 2\,e^{i 5\pi/6} = -\sqrt{3} + i$,
$w_{2} = 2\,e^{-i\pi/2} = -2i$. An estimator who sketches the
three corners of an equilateral triangle with one vertex on the
negative imaginary axis reads the angles to within better than
$0.1\,\text{rad}$ from the sketch. The sum $w_{0} + w_{1} + w_{2}
= 0$ as a check (the three cube roots of any nonzero complex
number sum to zero).
\end{exercise}

\begin{exercise}
Estimate the resonant frequency of a series RLC circuit with
$L = 1 \,\si{\milli\henry}$ and $C = 100 \,\si{\nano\farad}$,
without a calculator. State the formula $\omega_{0} = 1/\sqrt{LC}$
and the unit checks. Then compute and report the answer in
hertz.

\paragraph{Solution sketch.}
$L = 10^{-3}\,\si{\henry}$, $C = 10^{-7}\,\si{\farad}$, so $LC =
10^{-10}\,\si{\second\squared}$ and $\sqrt{LC} = 10^{-5}\,\si
{\second}$. Then $\omega_{0} = 1/\sqrt{LC} = 10^{5}\,\si{\radian
\per\second}$. In hertz, $f_{0} = \omega_{0}/(2\pi) = 10^{5}/6.283
\approx 1.59 \times 10^{4}\,\si{\hertz} \approx 15.9\,\si
{\kilo\hertz}$. Unit check: $\si{\henry} \cdot \si{\farad} =
(\si{\volt}\cdot\si{\second}/\si{\ampere}) \cdot (\si{\ampere}\cdot
\si{\second}/\si{\volt}) = \si{\second\squared}$, so $1/\sqrt{LC}$
is in $\si{\second^{-1}}$, that is, $\si{\radian\per\second}$,
as required.
\end{exercise}

\begin{exercise}
Estimate the phase shift across a $1 \,\si{\milli\farad}$
capacitor driven at $\omega = 2\pi \cdot 60 \,\si{\hertz}$ in a
circuit with series resistance $R = 1 \,\si{\ohm}$. State whether
the current leads or lags the voltage. Compute the exact phase
angle and compare.

\paragraph{Solution sketch.}
$\omega = 2\pi \cdot 60 \approx 377\,\si{\radian\per\second}$;
capacitive reactance $1/(\omega C) = 1/(377 \cdot 10^{-3}) \approx
2.65\,\si{\ohm}$. The capacitive reactance is about $2.65$ times
the resistance; the circuit is dominated by capacitance but not
overwhelmingly. The current leads the voltage (capacitor); the
phase angle of the impedance is in the fourth quadrant.
$\varphi_{Z} = \arctan(-2.65/1) \approx -1.21\,\text{rad} \approx
-69.4^{\circ}$. The current leads the voltage by about
$69.4^{\circ}$. An estimator who anchors on the reactance-to-
resistance ratio of $2.65$ would estimate near $-70^{\circ}$,
within a degree of the exact value.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
In a programming language of the reader's choice, plot $e^{i\theta}$
in the complex plane for $\theta \in [0, 2\pi]$ at $100$ equally
spaced values. Verify visually that the trace is the unit circle.
Repeat for $\theta \in [0, 6\pi]$ to verify the periodicity.

\paragraph{Solution sketch.}
In Python:
\begin{verbatim}
import numpy as np, matplotlib.pyplot as plt
theta = np.linspace(0, 2*np.pi, 100)
z = np.exp(1j*theta)
plt.plot(z.real, z.imag, '.'); plt.axis('equal'); plt.show()
\end{verbatim}
The hundred dots fall on the unit circle. For $\theta \in [0,
6\pi]$ the same hundred-point sampling shows three superposed
traces (the period is $2\pi$); the trace returns to its start at
$\theta = 2\pi$, again at $\theta = 4\pi$, and again at $\theta =
6\pi$. The exercise verifies, at the level of a picture, that
$e^{i(\theta + 2\pi k)} = e^{i\theta}$ for every integer $k$.
\end{exercise}

\begin{exercise}
Numerically integrate the second-order ODE for the series RLC
circuit with parameters of your choice (a reasonable starting
set is in the chapter project). Run for at least ten periods of
the drive frequency. Plot the source voltage and the current on
the same axes; identify the transient and the steady-state regimes.

\paragraph{Solution sketch.}
The reference implementation is in
\verb|volumes/02-form/ch03-trigonometry-complex-phasors/code/rlc_response.py|.
The first few periods show a transient: the current's amplitude
and phase oscillate before settling. The transient decays on the
time scale $\tau = 2L/R$ (for the underdamped second-order system);
the steady-state regime begins once the residual transient drops
below the precision of the comparison ($\sim 1\%$). The
steady-state current trace matches the phasor result $|\tilde I|
\cos(\omega t + \arg \tilde I)$ to within numerical-integration
error. The phasor analysis recovers the steady-state amplitude
and phase in three lines; the ODE recovers them by waiting for
the transient to die.
\end{exercise}

\begin{exercise}
Generate a sinusoidal phase signal $\theta(t) = \omega t$ for
$\omega = 2\pi$ and $t \in [0, 5]$. Wrap the signal to
$(-\pi, \pi]$ by computing $\theta_{\text{wrap}}(t) =
\operatorname{atan2}(\sin\theta(t), \cos\theta(t))$. Plot the
wrapped signal. Then
implement a simple unwrap algorithm that adds $2\pi$ whenever
the difference between consecutive samples exceeds $\pi$ in
magnitude. Plot the unwrapped signal and verify it matches the
original $\theta(t)$.

\paragraph{Solution sketch.}
The reference implementation is in
\verb|volumes/02-form/ch03-trigonometry-complex-phasors/code/phase_unwrap.py|.
The wrapped trace shows a sawtooth: linear ramp from $-\pi$ to
$+\pi$ over each period, then a jump back to $-\pi$. The unwrap
algorithm tracks a running cycle count $k$, increments or
decrements $k$ at every jump, and adds $2\pi k$ to each sample.
With $5001$ samples over $5$ seconds (step $1\,\si{\milli\second}$),
the sample-to-sample phase change is $\omega\,\Delta t \approx
0.006\,\text{rad}$, well below the $\pi$-radian threshold; the
unwrap recovers $\theta(t)$ within floating-point error. With
$9$ samples (step $5/8 = 0.625\,\si{\second}$), the
sample-to-sample change is $\omega\,\Delta t \approx 3.93\,\text{rad}
> \pi$; the unwrap fails silently.
\end{exercise}

\begin{exercise}
Simulate two waves $y_{1}(t) = \cos(2\pi t)$ and $y_{2}(t) =
\cos(2\pi t + \varphi)$ for $\varphi \in \{0, \pi/4, \pi/2,
3\pi/4, \pi\}$. For each value of $\varphi$, compute the
amplitude of the sum $y_{1}(t) + y_{2}(t)$ and verify it matches
the formula $\sqrt{2 + 2\cos\varphi}$.

\paragraph{Solution sketch.}
The expected amplitudes are $\sqrt{2 + 2\cos\varphi} = 2|\cos
(\varphi/2)|$ at $\varphi \in \{0, \pi/4, \pi/2, 3\pi/4, \pi\}$:
$2$, $\sqrt{2 + \sqrt{2}} \approx 1.848$, $\sqrt{2} \approx 1.414$,
$\sqrt{2 - \sqrt{2}} \approx 0.765$, $0$. Sample $y_{1}(t) +
y_{2}(t)$ on $t \in [0, 2]$ at one thousand points and read off
the maximum absolute value; agreement with the formula is to
better than $0.1\%$. The endpoints $\varphi = 0$ (constructive)
and $\varphi = \pi$ (destructive) recover the textbook limits.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a series RLC band-pass filter with centre frequency $1
\,\si{\kilo\hertz}$ and quality factor $Q = 10$. State the
component values for $L$ and $C$, and the resulting bandwidth.
Specify the resistor value $R$ from the relation $Q = \omega_{0}
L / R$. Identify which component sets the bandwidth.

\paragraph{Discussion.}
$\omega_{0} = 2\pi \cdot 1000 \approx 6283\,\si{\radian\per\second}$,
so $LC = 1/\omega_{0}^{2} \approx 2.53 \times 10^{-8}\,\si{\second
\squared}$. The values of $L$ and $C$ separately are free; common
practice is to pick $L$ in the millihenry range and back out $C$.
With $L = 10\,\si{\milli\henry}$, $C = 1/(\omega_{0}^{2} L) \approx
2.53\,\si{\micro\farad}$. The bandwidth is
$\Delta\omega = \omega_{0}/Q$, so $\Delta f = f_{0}/Q = 100\,\si
{\hertz}$. The resistor sets the bandwidth through $Q = \omega_{0}
L / R$: $R = \omega_{0} L / Q = 6283 \cdot 0.010 / 10 \approx
6.28\,\si{\ohm}$. Larger $R$ widens the bandwidth and lowers
selectivity; smaller $R$ narrows the bandwidth at the cost of
sharper ringing in the transient response. A practical design
includes the inductor's intrinsic series resistance in the
budget and trims the explicit resistor accordingly.
\end{exercise}

\begin{exercise}
Design a power-factor correction scheme for an industrial load
that draws apparent power $S = 100 \,\si{\kilo\text{VA}}$ at a
power factor of $0.7$ lagging at $50 \,\si{\hertz}$. State the
required reactive power compensation and the capacitance needed
in parallel with the load to bring the power factor to $0.95$
lagging. State the assumption that the load is at $400 \,\si
{\volt}$ line-to-line.

\paragraph{Discussion.}
Real power $P = S\cos\varphi = 100 \cdot 0.7 = 70\,\si{\kilo
\watt}$. Existing reactive power $Q = S\sin\varphi$ with
$\sin\varphi = \sqrt{1 - 0.49} \approx 0.714$, so $Q \approx
71.4\,\si{\kilo\text{var}}$. Target reactive power at $\cos\varphi'
= 0.95$: $Q' = P\tan\varphi' = 70 \cdot 0.329 \approx 23.0\,
\si{\kilo\text{var}}$. The capacitor must supply $\Delta Q = Q -
Q' \approx 48.4\,\si{\kilo\text{var}}$ of leading reactive power.
For a delta-connected three-phase capacitor bank at $400\,\si
{\volt}$ line-to-line, each capacitor sees the full
line-to-line voltage and supplies $\Delta Q/3$. Each leg's
capacitance is $C = (\Delta Q / 3) / (\omega V_{LL}^{2})
\approx 16{,}100 / (314 \cdot 160{,}000) \approx 3.2 \times
10^{-4}\,\si{\farad} = 320\,\si{\micro\farad}$. Three capacitors of
$320\,\si{\micro\farad}$ each, total $960\,\si{\micro\farad}$.
A wye-connected bank would need three capacitors of $3 \cdot
320 = 960\,\si{\micro\farad}$ each at the line-to-neutral voltage
$231\,\si{\volt}$. The design also requires a fused disconnect,
discharge resistors to bleed off stored charge after
disconnection, and protection against switching transients.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A measurement of phase angle in a synchrophasor record shows a
sudden $2\pi$ jump between two consecutive samples; the rest of
the record is smooth. State the most likely mechanism, propose
the next measurement to confirm it, and describe the unwrap
fix.

\paragraph{Solution sketch.}
The most likely mechanism is a wrap of the reported phase from
the principal range $(-\pi, \pi]$ to itself; the underlying
physical phase advanced past $\pi$ and the instrument's reporting
convention discontinuously mapped it to $-\pi$. To confirm,
inspect the magnitude trace at the same instant: a real
phase-step event (a fault, a switching transient) is accompanied
by a magnitude perturbation, while a wrap leaves the magnitude
smooth. The fix is to apply a first-order unwrap algorithm: add
$2\pi$ to all samples after the jump (or subtract, depending on
the direction of the jump) so consecutive samples differ by less
than $\pi$. A reliable implementation also checks the sample-to-
sample phase change against the expected $\omega\,\Delta t$ for
the reporting rate; a true wrap matches that prediction within
the measurement noise.
\end{exercise}

\begin{exercise}
A program computes the phase of a complex number using the
expression $\theta = \arctan(\Im(z) / \Re(z))$ rather than the
four-quadrant $\operatorname{atan2}(\Im(z), \Re(z))$. The program
returns wrong values for some inputs. State the inputs for which
the program is wrong, the size of the error, and the fix.

\paragraph{Solution sketch.}
$\arctan$ returns a value in $(-\pi/2, \pi/2)$ and depends only on
the ratio $\Im/\Re$. Two points $z = a + bi$ and $-z = -a - bi$
share the same ratio but lie on opposite sides of the origin;
$\arctan$ collapses them to a single angle. The program is wrong
whenever $\Re(z) < 0$, that is, for points in the second and
third quadrants. The size of the error is exactly $\pi$ radians
($180^{\circ}$). The fix is to replace $\arctan$ with the
four-quadrant $\operatorname{atan2}(\Im(z), \Re(z))$, which uses
the signs of both arguments and returns a value in $(-\pi, \pi]$.
The bug is silent at the language level: the program returns a
number; the number is just wrong by $\pi$. Test coverage that
includes points in every quadrant catches the failure
immediately.
\end{exercise}

\begin{exercise}
A series RLC circuit's measured current amplitude exceeds the
predicted phasor amplitude by a factor of two. State three
plausible mechanisms (resonance not accounted for, component
values different from labelled, source frequency different from
expected, nonlinear element such as a saturating inductor, the
measurement instrument operating outside its bandwidth) and
propose the next measurement that distinguishes them.

\paragraph{Solution sketch.}
\emph{Mechanism 1: resonance proximity}. The drive frequency is
close to $\omega_{0} = 1/\sqrt{LC}$, where the reactances cancel
and the impedance reduces to $R$ alone. Distinguishing
measurement: sweep the drive frequency by $\pm 10\%$ and observe
the current's response; near resonance, small frequency changes
produce large amplitude changes.
\emph{Mechanism 2: component value mismatch}. The labelled values
of $L$ or $C$ differ from actual by tolerance ($\pm 10\%$ for
low-grade parts, $\pm 1\%$ for precision parts). Distinguishing
measurement: characterise each component with an LCR meter at
the operating frequency.
\emph{Mechanism 3: source frequency mismatch}. The source's
actual frequency differs from the assumed value. Distinguishing
measurement: read the source's output frequency on a calibrated
counter; verify against the assumed value.
A single bench session that runs the frequency sweep first, then
swaps in characterised components if the sweep does not resolve
the issue, then verifies the source frequency, isolates the
mechanism in under an hour.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A laser interferometer measuring a slow vibration over many
hours reports a step change in displacement of $\sim 316 \,\si
{\nano\meter}$ at one instant; the underlying mechanical
vibration is smooth. Argue, in 200 to 300 words, the failure
mode that produced the apparent step, the operational discipline
that would have caught it before the measurement was reported,
and the redundant measurement that would have closed the gap.

\paragraph{Solution sketch.}
A laser interferometer at the helium-neon line $\lambda \approx
633\,\si{\nano\meter}$ has wrap distance $\lambda/2 \approx 316\,
\si{\nano\meter}$ (half a wavelength corresponds to one round-trip
fringe). The step change of exactly this magnitude is the
unmistakable signature of a missed wrap: the unwrap algorithm
failed to add a $2\pi$ phase to one sample, displacing the
recovered phase trajectory by half a wavelength. The discipline
that catches it is to choose the sampling interval so the
expected sample-to-sample displacement stays below $\lambda/4$
(quarter wavelength), and to flag any sample-to-sample step
larger than $\lambda/8$ for manual review. The discipline that
prevents it from corrupting the record is to verify the unwrap
end-to-end against an independent coarser sensor: a capacitive
displacement gauge, an LVDT, or a fibre-optic strain sensor
operating over the same span. A discrepancy of half a wavelength
between the interferometer and the coarse sensor is the
operational definition of a missed wrap; the discrepancy is
flagged before the measurement is reported and the unwrap is
re-run with the missed cycle inserted. Without the coarse
reference, the missed wrap propagates silently and corrupts
every downstream statistic.
\end{exercise}

\begin{exercise}
A radar Doppler processor reports a target velocity of $50 \,\si
{\meter\per\second}$ that is inconsistent with prior knowledge
(a static target). The pulse repetition frequency is $1 \,\si
{\kilo\hertz}$ and the wavelength is $3 \,\si{\centi\meter}$.
Argue, in 200 to 300 words, the wrap mechanism that may have
produced the spurious velocity, the unambiguous velocity range
implied by the pulse repetition frequency, and the operational
fix.

\paragraph{Solution sketch.}
The radar measures pulse-to-pulse phase change $\Delta\varphi$,
related to radial velocity by $\Delta\varphi = 2 \cdot 2\pi
v\,T_{p}/\lambda$, where $T_{p} = 1/\text{PRF} = 10^{-3}\,\si
{\second}$ is the pulse interval and the leading factor of two
accounts for the round trip. The unambiguous velocity range is
set by $|\Delta\varphi| \le \pi$: $v_{\max} = \lambda \cdot
\text{PRF}/4 = 0.03 \cdot 1000 / 4 = 7.5\,\si{\meter\per\second}$.
The reported velocity of $50\,\si{\meter\per\second}$ exceeds
$v_{\max}$ by a factor of $\sim 7$, so the report is consistent
with several integer-wrapped aliases of a true velocity. A
plausible true velocity in the unambiguous range is
$50 - n \cdot 15\,\si{\meter\per\second}$ for integer $n$;
$n = 3$ gives $v = 5\,\si{\meter\per\second}$, $n = 4$ gives
$v = -10\,\si{\meter\per\second}$, and $n = 0$ gives the original
reading. The operational fix is to use staggered pulse repetition
frequencies that disambiguate the wrap (a Chinese-remainder-theorem
construction recovers the true velocity from two staggered PRFs
whose unambiguous ranges are mutually prime). The discipline is
to select PRF based on the expected velocity range and to flag
any velocity report close to the wrap boundary for unambiguity
analysis.
\end{exercise}

\subsection*{Judgement}\pathtag{standard}

\begin{exercise}
A junior engineer proposes to analyse a sinusoidally driven
nonlinear circuit (a half-wave rectifier feeding a smoothing
capacitor) using phasor methods. Write a one-paragraph response
that takes the proposal seriously, identifies the regime in
which a phasor approximation could give working answers (the
small-signal regime around an operating point) and the regime
in which it cannot (the large-signal regime where the rectifier
nonlinearity dominates).

\paragraph{Discussion.}
Phasor analysis assumes a linear time-invariant system at a
single frequency. The half-wave rectifier is not linear:
the diode is conducting in one half of each cycle and blocking in
the other, and the smoothing capacitor's charge state depends on
the full waveform history. Direct phasor analysis at the source
frequency reports the resistive impedance of the diode-capacitor
network as if it were a linear element; the answer is wrong by
order unity. The small-signal regime where the proposal could
give working answers is a perturbation around an operating point:
once the rectifier reaches steady state with a fixed average
output voltage and a small superimposed ripple, the ripple itself
can be analysed by treating the diode's incremental conductance
as a linear resistance and applying phasor methods to the small
signal. The reader who wants the full waveform (rectified
fundamental, harmonics at the line frequency and its multiples,
the DC component) must use Fourier analysis or a time-domain
SPICE simulation; phasor analysis at the line frequency alone
recovers only a fraction of the answer. The right response to
the junior engineer is not to dismiss the proposal but to draw
the boundary: phasor analysis for the small-signal ripple,
Fourier or time-domain for the full nonlinear response.
\end{exercise}

\begin{exercise}
Argue, in 200 to 300 words, when phasor analysis is the
appropriate working tool for a steady-state AC system and when
the engineer should reach for the full time-domain integration
or for a numerical simulation instead. Take a position for one
class of system (linear LTI at a single frequency) where phasor
analysis is the right choice, and one class (nonlinear, transient,
multi-frequency, parametric) where it is not.

\paragraph{Discussion.}
Phasor analysis is the right tool when three conditions hold:
the system is linear (superposition applies, and the response to
a sum is the sum of responses); the system is time-invariant (the
impulse response does not depend on absolute time); and the input
is a single sinusoid in steady state (the transient has decayed).
A grid-connected synchronous motor at nominal voltage and
frequency satisfies all three conditions over time scales of
seconds to minutes; the phasor calculation gives the operating
point in three lines. A switched-mode power supply running on a
distorted mains voltage with harmonics into the seventh order
satisfies the first two conditions but not the third; the phasor
calculation at the fundamental misses the harmonic power
explicitly and gives the wrong total dissipation. A
parametrically-pumped oscillator (a Mathieu equation; the
parameter varies at twice the natural frequency) satisfies
neither time invariance nor the LTI conditions and is wholly
outside phasor analysis. Time-domain integration recovers all
three classes; phasor analysis recovers the first one fast and
the second one only as the leading-order approximation. The
discipline is to ask which conditions the system meets before
reaching for the tool, not after.
\end{exercise}

\begin{exercise}
A textbook describes a balanced three-phase voltage system as
the sum of three sinusoids. Argue, in 200 to 300 words, why the
phasor representation is the right working tool for analysing
balanced and unbalanced three-phase systems, what the symmetric-
components decomposition (positive, negative, and zero
sequences) buys, and where phasor analysis stops being adequate
(during a fault, during a switching transient, in the presence
of substantial harmonics).

\paragraph{Discussion.}
Three-phase phasors compress nine real numbers (three voltages
times three signed magnitudes-and-phases) into three complex
numbers. A balanced system has phasors related by $e^{\pm i 2\pi/3}$
rotations; the per-phase calculation reduces to a single-phase
calculation against a virtual neutral. The symmetric-components
decomposition introduced by Fortescue (1918) factors any
three-phase set into a positive-sequence (rotation by
$e^{-i 2\pi/3}$), a negative-sequence (rotation by $e^{+i 2\pi/3}$),
and a zero-sequence (all three components equal). A balanced
system has only positive-sequence content; a single-phase fault
excites all three sequences in known ratios; a phase imbalance
generates negative-sequence content that heats induction-motor
rotors and limits derating. The sequence decomposition is a
linear change of basis on the three-dimensional phasor space and
preserves the phasor algebra; the engineer who analyses a fault
in sequence components writes three independent single-phase
calculations rather than one coupled three-phase calculation.
Phasor analysis stops being adequate when the system is far from
its steady state (during the first few cycles after a switching
event, during the breaker operation of a fault clearance, or in
the presence of harmonics generated by power electronics). The
contemporary substation runs phasor-based protection at the
fundamental and supplements it with time-domain travelling-wave
analysis for fast-transient location \cite{paper:v2ch03-itrich-c37-118}.
\end{exercise}
