\chapter{Units and dimensions}
\label{vol01:ch02}

\epigraph{Define the terms before trusting the argument.}{Engineering
keeps the old philosophical demand and adds the engineering one:
define the units.}

\chapmeta{Half-life: foundational. Archetypes: balance (the first
balance archetype, in dimensional form). Exercise target: 35. Project
track: analysis. Hazard class: none. Reader path default: standard,
with sections explicitly tagged. Named cases: Mars Climate Orbiter
(1999), Air Canada Flight 143 (1983), and Ariane 5 Flight 501 (1996),
briefly revisited from chapter 1; Hubble Space Telescope primary
mirror (1990), as a metrology-chain failure; Therac-25 (1985--1987),
only where display quantities and operator interpretation are at
issue.}

\noindent
A measurement becomes usable only when its number, unit, and
dimensional signature travel together. This chapter trains that
discipline: carry the unit, check the dimensions, and treat every
unitless number in an engineering calculation as suspect until it is
explained.

\section{SI base units}\pathtag{core}

The International System of Units, in the form adopted by the General
Conference on Weights and Measures in 2018 and effective from 20 May
2019, has seven base units defined by fixed numerical values of seven
defining constants \cite{std:bipm-si2019}. The redefinition matters
for engineering because every base unit is now invariant: it is
anchored to a fixed numerical value, not to an artifact that can drift
or be lost.

The seven defining constants:

\begin{itemize}[leftmargin=*]
  \item The unperturbed ground-state hyperfine transition frequency of
    the caesium-133 atom, $\Delta\nu_{\text{Cs}} = 9{,}192{,}631{,}770
    \,\si{\hertz}$, defines the \engterm{second}.
  \item The speed of light in vacuum, $c = 299{,}792{,}458 \,\si{\meter
    \per\second}$, defines the \engterm{metre}.
  \item The Planck constant, $h = 6.626\,070\,15 \times 10^{-34}
    \,\si{\joule\second}$, defines the \engterm{kilogram}.
  \item The elementary charge, $e = 1.602\,176\,634 \times 10^{-19}
    \,\si{\coulomb}$, defines the \engterm{ampere}.
  \item The Boltzmann constant, $k_{\text{B}} = 1.380\,649 \times
    10^{-23} \,\si{\joule\per\kelvin}$, defines the \engterm{kelvin}.
  \item The Avogadro constant, $N_{\text{A}} = 6.022\,140\,76 \times
    10^{23} \,\si{\per\mole}$, defines the \engterm{mole}.
  \item The luminous efficacy of monochromatic radiation of frequency
    $540 \times 10^{12} \,\si{\hertz}$, $K_{\text{cd}} = 683 \,\si
    {\lumen\per\watt}$, defines the \engterm{candela}.
\end{itemize}

\input{volumes/01-quantity/ch02-units-and-dimensions/figures/fig-si-base-constants.tex}

The seven constants and the seven units form a closed loop: each
constant is defined exactly, each unit is then derived from the
constants without further measurement. The kilogram, for example, is
no longer the mass of a platinum-iridium cylinder kept in a vault near
Paris. It is whatever mass corresponds to a Planck constant of
$h = 6.626\,070\,15 \times 10^{-34} \,\si{\joule\second}$, with
$\si{\joule} = \si{\kilogram\meter\squared\per\second\squared}$. A
laboratory realises the kilogram with a Kibble balance or an
X-ray-crystal-density measurement of a silicon sphere; either method
is traceable, in principle, to the defining constant alone
\cite[ch.~2]{std:bipm-si2019}.

This matters because artifacts drift. The international prototype of
the kilogram, the platinum-iridium cylinder used as the reference from
1889 to 2019, diverged from its sister copies over repeated
verifications. BIPM reports a median difference of about $25 \,\si
{\micro\gram}$ for the original prototypes in the 1989--1991
verification; NIST summaries describe drifts of order
$50 \,\si{\micro\gram}$ over a century for official copies
\cite{std:bipm-si2019}. The direction could not be settled from the
artifact system alone, because the prototype \emph{was} the
definition. The 2019 redefinition cut that knot.

The candela is the base unit whose definition most visibly carries a
human observer. Its fixed constant is a luminous efficacy at one
frequency, so photometry can report lighting, displays, and visual
response in a common unit without pretending that visual brightness
is radiant power alone.

A reader who finishes this chapter should be able to recite the seven
defining constants with their numerical values and units. We do not
require the reader to memorise all the digits. We do require the
reader to know that the defining constants are exact by definition,
that any subsequent improvement in the realisation of an SI unit
sharpens what the constants \emph{realise} but never changes what
they \emph{are}, and that this design choice is what lets a
calibration in S\~ao Paulo agree with one in Stockholm to many
decimal places.

\subsection{From artefacts to constants: a working history}
\pathtag{standard}

The seven-constant SI of 2019 is the endpoint of a two-century arc.
The arc is worth walking briefly, because every base unit on the
present list passed through at least one artefact-anchored definition,
and the engineering lesson of the 2019 redefinition is that an
artefact is a metrological liability the moment one cares about
parts in $10^{8}$.

\paragraph{1799: the metre as a quarter-meridian artefact.} The
French Republic's commission of weights and measures, charged in
1791 with rationalising French metrology, defined the metre as one
ten-millionth of the quarter meridian from the equator to the North
Pole through Paris. The triangulation that fixed the length, carried
out by Pierre M\'echain and Jean-Baptiste Delambre between 1792 and
1798, supplied the number for the platinum bar that became the
\emph{m\`etre des Archives}, deposited on 22 June 1799 in the
French national archives \cite{hist:ch02t3-prototype-metre-1799}.
The artefact, not the meridian arc, became the definition the
moment the bar was certified. Subsequent measurements of the
meridian found the arc to differ from the bar by about
$0.2 \,\si{\milli\meter}$ in $1 \,\si{\meter}$; the bar was kept.
The lesson was already audible: once a physical object is the
reference, the geometric definition is a story about its origin,
not a check on its value.

\paragraph{1889: the international prototype kilogram (IPK).} The
First General Conference on Weights and Measures (CGPM) in 1889
sanctioned a platinum-iridium cylinder (90\% Pt, 10\% Ir) as the
international prototype of the kilogram, deposited at the Bureau
International des Poids et Mesures (BIPM) at S\`evres
\cite{std:bipm-si2019}. Forty national prototypes were manufactured
to within a few tens of micrograms of the international prototype.
Periodic comparisons over the next century, in 1889--1911,
1939--1953, and 1988--1992, found the national prototypes drifting
relative to the international prototype by approximately $50 \,\si
{\micro\gram}$ over the century, in one direction or another,
without a way to decide whether the international prototype or the
national prototypes had moved
\cite{web:bipm-kilogram-history}. The discipline was forced into
a metrology in which the reference \emph{was} the definition; the
artefact's drift could be measured only against itself.

\paragraph{1960: the metre via krypton-86.} The 11th CGPM in 1960
adopted a quantum-mechanical definition of the metre as
$1{,}650{,}763.73$ wavelengths in vacuum of the radiation
corresponding to the $2p_{10} \to 5d_{5}$ transition of the
krypton-86 atom \cite{web:ch02t3-bipm-cgpm-1960}. The platinum-
iridium prototype bar was retired as the metre's reference and
became a museum object. The redefinition reduced the metre's
uncertainty by roughly two orders of magnitude (from a few parts in
$10^{7}$ for the bar to a few parts in $10^{9}$ for the wavelength
realisation) and removed the dependence on a single artefact. The
choice of $1{,}650{,}763.73$ was set so that the new metre matched
the old metre at the third decimal of a micrometre.

\paragraph{1967: the second via caesium-133.} The 13th CGPM in
1967/68 defined the second as the duration of $9{,}192{,}631{,}770$
periods of the radiation corresponding to the transition between
the two hyperfine levels of the caesium-133 ground state
\cite{web:ch02t3-bipm-cgpm-1967}. The redefinition replaced the
ephemeris second (a fraction of the tropical year 1900) with an
atomic-transition standard whose fractional uncertainty at the time
of adoption was roughly $10^{-12}$ and is now (in the best
caesium-fountain primary standards as of 2026) about $10^{-16}$
\cite{web:nist-cesium-fountain}. The second became the first base
unit defined by a fixed numerical value of a frequency; the
template for the 2019 SI was set.

\paragraph{1983: the metre via the speed of light.} The 17th CGPM
in 1983 redefined the metre as the length of the path travelled by
light in vacuum during a time interval of
$1/299{,}792{,}458 \,\si{\second}$ \cite{web:ch02t3-bipm-cgpm-1983}.
The speed of light became exact by definition; the metre became
derived from the second through the speed of light. The
krypton-86 wavelength definition of 1960 was retired. The 1983
redefinition is the first in which a base unit was defined by
fixing the numerical value of a fundamental constant. The same
move, generalised, would arrive in 2019 for four more base units.

\paragraph{2018/2019: the four-constant redefinition.} The 26th
CGPM, meeting on 16 November 2018, adopted Resolution 1 redefining
the kilogram, the ampere, the kelvin, and the mole in terms of
fixed numerical values of, respectively, the Planck constant $h$,
the elementary charge $e$, the Boltzmann constant $k_{\text{B}}$,
and the Avogadro constant $N_{\text{A}}$
\cite{web:ch02t3-bipm-comptes-rendus-2018, paper:stock-kibble-2018}.
The redefinitions took effect on World Metrology Day, 20 May 2019.
The international prototype kilogram of 1889 was retired after 130
years of service; its mass is now reported as the result of a
measurement, like any other kilogram artefact.

\paragraph{Why the kilogram redefinition mattered for working
engineers, and why it did not.} The realisations of the kilogram
through the Kibble balance and the silicon-sphere X-ray-crystal-
density method now have combined relative uncertainties at the
$\sim 1 \times 10^{-8}$ level (as of 2026)
\cite{web:bipm-mise-en-pratique-kilogram-2019,
web:nist-kibble-balance}. The continuity requirement built into
the 2018 resolution was that the new kilogram should match the old
kilogram to better than the uncertainty of the realisation, so
that no industrial scale, balance, or trade mass needed
recalibration on the night of 19--20 May 2019. The redefinition
mattered, in working terms, only for the few national metrology
institutes whose Kibble balances and silicon-sphere apparatus were
being recalibrated against the new definition; the
\emph{mise en pratique} for the kilogram, published as a BIPM CCM
document, gives the realisation procedures
\cite{web:bipm-mise-en-pratique-kilogram-2019}. For the engineer
weighing a part on a calibrated lab balance, the practical effect
of the redefinition is invisible. The conceptual effect is large:
the kilogram is now an attribute of nature, not of a vault.

\section{Derived units}\pathtag{core}

Every other unit in SI is built from products and ratios of the seven
base units. The system is \engterm{coherent}: a derived unit is
exactly the corresponding combination of base units, with no extra
numerical factor. If a quantity has the dimensions of mass times
length per time squared, the SI unit is the kilogram metre per second
squared, which is named the newton. There is no conversion factor
between $\si{\newton}$ and $\si{\kilogram\meter\per\second\squared}$;
they are the same unit under two names \cite{std:bipm-si2019}.

The named coherent derived units the reader will use most often:

\begin{itemize}[leftmargin=*]
  \item $\si{\hertz} = 1/\si{\second}$: frequency.
  \item $\si{\radian}$: plane angle, dimensionless.
  \item $\si{\steradian}$: solid angle, dimensionless.
  \item $\si{\newton} = \si{\kilogram\meter\per\second\squared}$:
    force.
  \item $\si{\pascal} = \si{\newton\per\meter\squared}$: pressure.
  \item $\si{\joule} = \si{\newton\meter}$: energy.
  \item $\si{\watt} = \si{\joule\per\second}$: power.
  \item $\si{\coulomb} = \si{\ampere\second}$: electric charge.
  \item $\si{\volt} = \si{\joule\per\coulomb} = \si{\watt\per
    \ampere}$: electric potential.
  \item $\si{\ohm} = \si{\volt\per\ampere}$: resistance.
  \item $\si{\farad} = \si{\coulomb\per\volt}$: capacitance.
  \item $\si{\henry} = \si{\volt\second\per\ampere}$: inductance.
  \item $\si{\tesla} = \si{\volt\second\per\meter\squared}$: magnetic
    flux density.
  \item $\si{\lumen} = \si{\candela\steradian}$: luminous flux.
  \item $\si{\lux} = \si{\lumen\per\meter\squared}$: illuminance.
\end{itemize}

The full list of 22 named coherent derived units is in the SI
brochure \cite[ch.~3]{std:bipm-si2019}. The reader should treat the
list as engineering vocabulary: a quantity expressed in joules per
kilogram is specific energy; one expressed in pascal seconds is
dynamic viscosity; one expressed in newton metres is either torque or
energy depending on context, and the convention is that torque is
written $\si{\newton\meter}$ rather than $\si{\joule}$ even though
the units are identical.

A few non-SI units are accepted for use with SI because they are
unavoidable in practice or because they fit cleanly into engineering
work: minute, hour, day, degree of arc, hectare, litre, tonne (also
called the metric ton), electronvolt, dalton (formerly the atomic
mass unit), and the astronomical unit. These are not coherent: a
litre is exactly $10^{-3} \,\si{\meter\cubed}$, a tonne is exactly
$10^{3} \,\si{\kilogram}$, an electronvolt is exactly
$1.602\,176\,634 \times 10^{-19} \,\si{\joule}$ in the revised SI.
The reader uses them for legibility, not for calculation; in any
expression where units must cancel, convert to coherent units first.

The decimal prefixes form a parallel system, attached to either base
or derived units, scaling by powers of ten. The prefixes go from
$10^{-30}$ (quecto) to $10^{30}$ (quetta); the 27th General Conference
added ronna, quetta, ronto, and quecto in 2022, after the 2019
redefinition \cite{std:bipm-si2019}. The prefixes that appear
constantly in engineering are: kilo $10^{3}$, mega $10^{6}$, giga
$10^{9}$, tera $10^{12}$, milli $10^{-3}$, micro $10^{-6}$, nano
$10^{-9}$, and pico $10^{-12}$. A reader who learns these eight has
covered most of working life.

A small caution. The prefix and the unit are written without a space:
$\si{\kilo\watt}$, not ``kilo watt''; $\si{\milli\meter}$, not ``milli
metre.'' The exception is the unit symbol itself, which is separated
from the number by a non-breaking thin space: $5 \,\si{\meter}$, not
``$5\si{\meter}$.'' Typesetting engines such as \texttt{siunitx} apply
this convention automatically, and the source files of this book do
the same. In hand-written work, the reader is expected to maintain it
without the engine's help.

\section{Conversion and the discipline of carrying units}\pathtag{core}

A measured quantity is a number paired with a unit. Neither half is
optional. The quantity is invariant; the number depends on the unit.
A length of $1$ metre is also $100$ centimetres, also $1000$
millimetres, also $39.37$ inches, also $1.094$ yards, all describing
the same length. The number changes; the length does not.

Write the unit on each quantity as the calculation moves. That habit
checks dimensions while the work is still live, makes conversions
explicit, and lets a six-month-old notebook remain auditable.

A conversion is multiplication by an expression equal to one. The
expression is a ratio of two quantities that are equal: $1\,\text{ft}
= 0.3048 \,\si{\meter}$ exactly, so

\[
1 = \frac{0.3048 \,\si{\meter}}{1 \,\text{ft}} = \frac{1 \,\text{ft}}
{0.3048 \,\si{\meter}}.
\]

To convert a foot to metres, write the original quantity, multiply by
the version of $1$ that cancels the unwanted unit, and let the
arithmetic run:

\[
1 \,\text{ft} \times \frac{0.3048 \,\si{\meter}}{1 \,\text{ft}} =
0.3048 \,\si{\meter}.
\]

The foot cancels because it appears once in the numerator and once in
the denominator. The metre is what is left. The same machine handles
every conversion we will ever need:

\[
20 \,\text{mph} \times \frac{1609.344 \,\si{\meter}}{1 \,\text{mile}}
\times \frac{1 \,\si{\hour}}{3600 \,\si{\second}} = 8.94 \,\si
{\meter\per\second}.
\]

The reader will see this discipline applied in every calculation in
the rest of the book. Equations carry their units. Conversions appear
in writing. Numbers without units are debugging targets, not
measurements.

\subsection{Special cases}

A few conversions are not pure multiplication.

\engterm{Temperature scales} other than the kelvin involve an offset
as well as a scale factor. A change of one kelvin is a change of one
degree Celsius, but a temperature in kelvin and the same temperature
in Celsius differ by $273.15$:

\[
T_{\si{\degreeCelsius}} = T_{\si{\kelvin}} - 273.15.
\]

The Fahrenheit scale uses both a different offset and a different
scale factor:

\[
T_{{}^{\circ}\mathrm{F}} = \frac{9}{5} T_{{}^{\circ}\mathrm{C}} + 32.
\]

A reader who treats temperature conversions as pure multiplication
will produce wrong answers in the third digit. A common engineering
mistake is to convert a temperature \emph{difference} as if it were
an absolute temperature; for differences, only the scale factor
applies.

\engterm{Gauge pressure} and \engterm{absolute pressure} differ by
the local atmospheric pressure. The standard atmosphere is defined
exactly as $1\,\text{atm} = 101{,}325 \,\si{\pascal}$; local sea-level
pressure is often close to that value, but is a weather-dependent
measurement, not a definition. A car tyre inflated to
$32\,\text{psi}$ gauge is at about
$220{,}600 \,\si{\pascal}$ gauge, or $321{,}900 \,\si{\pascal}$
absolute. Engineering codes for vacuum, hydraulics, gas-handling, and
medicine specify which is meant; the reader who gets it wrong gets
either a tyre that explodes or a vacuum chamber that does not pump
down.

\engterm{The decibel} is a dimensionless non-SI unit accepted for
logarithmic ratio quantities. A sound-pressure level in
$\si{\deci\bel}$ is $20 \log_{10}(p/p_{0})$ where
$p_{0} = 20 \,\si{\micro\pascal}$. A voltage ratio uses
$20 \log_{10}(V/V_{0})$ when the impedance basis is fixed. A power
ratio uses $10 \log_{10}(P/P_{0})$, with the factor of $10$ rather
than $20$ because power scales as the square of root-power
quantities. We return to decibels in chapter 6 of this volume when we
treat signals.

\engterm{The pH scale} measures hydrogen-ion activity logarithmically:
$\mathrm{pH} = -\log_{10}(a_{\mathrm{H}^{+}})$. A pH of $5$ is ten
times more acidic than a pH of $6$, not slightly more. Logarithmic
scales reward the reader who distinguishes the underlying quantity
from its scaled representation.

\section{Dimensional analysis as debugging}\pathtag{core}

Every physical equation balances dimensions on both sides. This is
not a convention. It is a consequence of the fact that the laws of
physics do not depend on the choice of unit system. If we write a
correct equation in SI and then convert every unit in it to imperial,
the equation must remain correct. The only way that can happen is if
the dimensions on each side match, and the conversions on each side
balance.

We will name the seven independent base dimensions with bracketed
symbols: $[L]$ for length, $[M]$ for mass, $[T]$ for time, $[I]$ for
electric current, $[\Theta]$ for thermodynamic temperature, $[N]$ for
amount of substance, and $[J]$ for luminous intensity. Every quantity
in this book has a dimensional signature in these symbols. Force is
$[M][L][T]^{-2}$; energy is $[M][L]^{2}[T]^{-2}$; pressure is
$[M][L]^{-1}[T]^{-2}$; entropy is $[M][L]^{2}[T]^{-2}[\Theta]^{-1}$;
electrical conductivity is $[M]^{-1}[L]^{-3}[T]^{3}[I]^{2}$.

\begin{archetype}[Balance, dimensional form]
The first archetype of this volume. Every term in a physical equation
has the same dimensional signature. If a sum has terms with different
signatures, at least one is wrong. If the left and right sides have
different signatures, the equation is wrong before any number is
computed. Dimensional balance is the cheapest, fastest, and most
reliable check an engineer has.
\end{archetype}

\subsection{The discipline of checking}

The check has two steps. The first is to write down the dimensions of
every quantity in the equation. The second is to verify that every
term has the same dimensions.

For Newton's second law, $F = ma$:

\[
[F] = [M][L][T]^{-2}, \qquad
[m][a] = [M] \times [L][T]^{-2} = [M][L][T]^{-2}.
\]

The dimensions match. We have not proved Newton's second law; we have
verified that whatever derivation produced it did not violate
dimensional consistency.

For the kinetic energy of a particle, $E = \tfrac{1}{2} m v^{2}$:

\[
[E] = [M][L]^{2}[T]^{-2}, \qquad
[m][v]^{2} = [M] \times ([L][T]^{-1})^{2} = [M][L]^{2}[T]^{-2}.
\]

The dimensions match. The factor of $\tfrac{1}{2}$ does not appear in
the dimensional check, because pure numbers (and dimensionless
factors generally) have no dimensional weight.

For the simple-pendulum period, $T = 2\pi \sqrt{L/g}$:

\[
[T] \stackrel{?}{=} \sqrt{[L]/[L][T]^{-2}} = \sqrt{[T]^{2}} = [T].
\]

The dimensions match. Dimensional analysis cannot tell us where the
$2\pi$ comes from. It can tell us that no sensible derivation will
give us a pendulum period of $\sqrt{L/(mg)}$, because that has the
wrong dimensions, or of $\sqrt{Lg}$, which also has the wrong
dimensions.

\begin{estimation}
\textbf{The terminal velocity of a small raindrop.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A raindrop of diameter $D = 2 \,\si{\milli\meter}$ falls through air.
At terminal velocity, the gravitational force on the drop is balanced
by the aerodynamic drag. Working dimensionally, what does the
terminal velocity scale as?

The variables that matter: the drop diameter $D$ ($[L]$), the drop
density $\rho_{w} \approx 10^{3} \,\si{\kilogram\per\meter\cubed}$
($[M][L]^{-3}$), the air density $\rho_{a} \approx 1.2 \,\si
{\kilogram\per\meter\cubed}$, the gravitational acceleration $g
\approx 9.8 \,\si{\meter\per\second\squared}$ ($[L][T]^{-2}$), and
the dynamic viscosity of air $\mu_{a} \approx 1.8 \times 10^{-5}
\,\si{\pascal\second}$. We want the terminal velocity $v$ ($[L][T]^
{-1}$).

For a raindrop of this size, the drag is mostly inertial (Reynolds
number is high), so $\mu_{a}$ drops out and we expect $v$ to be set
by a balance between gravity and inertial drag:

\[
\rho_{w} g D^{3} \sim \rho_{a} v^{2} D^{2},
\]

which gives

\[
v \sim \sqrt{\frac{\rho_{w}}{\rho_{a}} g D} \approx \sqrt{\frac{10^{3}}
{1.2} \times 9.8 \times 2 \times 10^{-3}} \approx \sqrt{16} \approx
4 \,\si{\meter\per\second}.
\]

The published value for a $2 \,\si{\milli\meter}$ raindrop is
approximately $6.5 \,\si{\meter\per\second}$
\cite{paper:gunn-kinzer1949}. The dimensional argument is low by
about a factor of $1.6$, which is well inside the factor-of-three
band a useful estimate is supposed to land in. The factor-of-three
check passed.

The estimate misses the prefactor because dimensional analysis cannot
recover prefactors; it also misses the buoyancy term and the
Reynolds-number-dependent drag coefficient that the dimensional
argument absorbs into its unknown function. Both refinements are
subjects of chapter 12 of Volume III.
\end{estimation}

\begin{estimation}
\textbf{The fluid-mechanical regime of a swimming bacterium.}

\smallskip
\textit{Estimate the Reynolds number for an \emph{E. coli} cell
swimming through water, before reading on.}

\smallskip
\emph{E. coli} is approximately $L \approx 2 \,\si{\micro\meter}$
long and swims at $v \approx 30 \,\si{\micro\meter\per\second}$.
Water density and viscosity at room temperature: $\rho = 10^{3}
\,\si{\kilogram\per\meter\cubed}$, $\mu = 10^{-3} \,\si{\pascal
\second}$. Reynolds number:

\[
\mathrm{Re} = \frac{\rho v L}{\mu} = \frac{10^{3} \times 30 \times
10^{-6} \times 2 \times 10^{-6}}{10^{-3}} = 6 \times 10^{-5}.
\]

The published value is consistent with $\mathrm{Re}$ in the range
$10^{-5}$ to $10^{-4}$ for swimming microbes \cite{paper:purcell-1977}.
The factor-of-three check passed.

The dimensional interpretation: bacteria swim in the Stokes-flow
regime, where viscous forces overwhelm inertia. The consequences are
striking. A bacterium that stops its flagellum coasts approximately
$0.1$ nanometres before viscous drag halts it; in inertial-dominated
fluid, a swimming body coasts substantial fractions of its body
length. Bacterial propulsion strategies (Purcell's ``Life at low
Reynolds number'' lecture is the canonical reference) must exploit
non-reciprocal motions to make net progress; reciprocal motions
(opening and closing a clam shell) produce zero net displacement in
Stokes flow. The dimensional regime determines the achievable
locomotion strategy.
\end{estimation}

\subsection{What dimensional analysis cannot do}

Dimensional analysis tells the reader the form of the answer. It does
not tell the reader the prefactor. The pendulum's $2\pi$ comes from
solving the harmonic-oscillator differential equation; the
raindrop's actual coefficient comes from measurements of the drag
coefficient as a function of Reynolds number; the $\tfrac{1}{2}$ in
kinetic energy comes from integrating $F \cdot v \,\dd t$ from rest.
Dimensional analysis is the structure of the answer; the prefactor
requires more work.

A second limit: dimensional analysis cannot detect errors that
preserve dimensions but use the wrong physics. An equation with the
right dimensions can still be wrong if its terms describe the wrong
mechanism, if it omits a term that should be included, or if its
constants come from a different regime. A successful dimensional
check is necessary, not sufficient.

A third limit: dimensional analysis cannot distinguish between
quantities that share the same dimensions. Torque and energy both
have dimensions $[M][L]^{2}[T]^{-2}$; they are physically distinct
and combine differently with other quantities. The discipline of
naming quantities and tracking their meaning, beyond their dimensional
signature, is what stops dimensional analysis from being a
license to multiply mismatched quantities.

\section{Dimensional homogeneity and the Buckingham pi
theorem}\pathtag{core}

The same check becomes a design tool in the \engterm{Buckingham pi
theorem}. For a dimensionally homogeneous relationship involving $n$
variables and a dimensional matrix of rank $k$, the relationship can
be written in terms of $n - k$ independent dimensionless groups.
Buckingham published the theorem in 1914 \cite{paper:buckingham1914};
Bridgman later systematised dimensional analysis for engineers and
physicists \cite{text:bridgman1922}; Barenblatt gives a modern
treatment of scaling and similarity \cite{text:barenblatt1996}.

\subsection{The theorem}

\begin{theorem}[Buckingham pi]
Let a physical relationship involve $n$ variables $q_{1}, q_{2},
\ldots, q_{n}$, where the dimensional matrix of these variables has
rank $k$ over the chosen base dimensions. Then the relationship can
be expressed as a function $f(\pi_{1}, \pi_{2}, \ldots, \pi_{n - k})
= 0$, where each $\pi_{j}$ is a dimensionless product of the
variables.
\end{theorem}

The proof is short; we defer it to the mathematical-modelling
treatment in Volume II and use the engineering procedure here.

\subsection{The procedure}

The reader follows these steps to apply the theorem to a problem.

\begin{enumerate}[leftmargin=*]
  \item List every variable that the problem might depend on. This is
    the step that requires physical judgement; an omitted variable
    will produce a wrong answer.
  \item Write each variable's dimensional signature in the chosen
    base dimensions.
  \item Count the number of variables, $n$, and the number of
    independent base dimensions actually appearing in the problem,
    $k$. The number of dimensionless groups is $n - k$.
  \item Choose $k$ of the variables as ``repeating variables.'' These
    appear in every dimensionless group. The repeating variables must
    together span the $k$ base dimensions; they should usually be the
    most basic geometric, kinematic, or material variables.
  \item Form each dimensionless group $\pi_{j}$ by combining the
    $k$ repeating variables with one of the remaining variables, and
    solving for the exponents that make the group dimensionless.
  \item Write the relationship as $\pi_{1} = \Phi(\pi_{2}, \ldots,
    \pi_{n-k})$ for some unknown function $\Phi$.
\end{enumerate}

\subsection{The named dimensionless groups}

A practising dimensional analyst recognises and names the working
dimensionless groups. The following are the ones the reader will
meet repeatedly through the chapters that follow. Each is a ratio
of physical quantities whose value determines a regime of
behaviour.

\engterm{Reynolds number} $\mathrm{Re} = \rho v L/\mu$ (inertia to
viscous forces in a flow). Transition from laminar to turbulent
pipe flow occurs around $\mathrm{Re} \approx 2300$ in the working
Moody-diagram convention; transition for flow over a flat plate
occurs around $\mathrm{Re}_{x} \approx 5 \times 10^{5}$. Volume IV
chapters $11$-$13$ work in Reynolds.

\engterm{Froude number} $\mathrm{Fr} = v/\sqrt{gL}$ (inertia to
gravity in free-surface flows). Naval architects use Froude scaling
for ship resistance: a $1{:}50$ scale-model ship in a towing tank
at Froude-matched speed $v_{m} = v_{p}/\sqrt{50}$ produces the
same wave pattern as the full-scale prototype. The transition from
sub-critical to super-critical open-channel flow occurs at
$\mathrm{Fr} = 1$.

\engterm{Mach number} $\mathrm{Ma} = v/c_{s}$ (flow velocity to
sound speed). Compressibility effects become important above
$\mathrm{Ma} \approx 0.3$; transonic effects appear around
$\mathrm{Ma} = 0.8$ to $1.2$; the shock-wave structure changes
between subsonic ($\mathrm{Ma} < 1$), sonic ($\mathrm{Ma} = 1$),
supersonic ($1 < \mathrm{Ma} < 5$), and hypersonic
($\mathrm{Ma} > 5$). Volume IV chapter $14$ develops the aerodynamic
applications.

\engterm{Weber number} $\mathrm{We} = \rho v^{2} L/\sigma$ (inertia
to surface tension). Drop breakup in a spray occurs when
$\mathrm{We}$ exceeds $\sim 12$ for a falling drop in still air;
inkjet printing tunes droplet ejection between $\mathrm{We} \sim
10$ and $100$ to produce satellite-free single drops. The
$\rho v^{2} L$ scaling makes Weber relevant when surface tension is
the dominant restoring force.

\engterm{Strouhal number} $\mathrm{St} = f L/v$ (oscillation
frequency, geometric scale, flow velocity). For vortex shedding
behind a circular cylinder, $\mathrm{St} \approx 0.21$ over a
wide Reynolds range ($\mathrm{Re} = 10^{2}$ to $10^{5}$); this is
the source of the Aeolian tone heard from wind across a stretched
wire (a $1\,\si{\milli\meter}$ wire in a $10\,\si{\meter\per\second}$
wind produces $\sim 2.1\,\si{\kilo\hertz}$, audible). The Tacoma
Narrows Bridge resonance (1940) involved Strouhal-driven aeroelastic
flutter.

\engterm{Prandtl number} $\mathrm{Pr} = \mu c_{p}/k$ (momentum to
thermal diffusivity). For air at room temperature, $\mathrm{Pr}
\approx 0.71$; for water, $\mathrm{Pr} \approx 7$; for liquid
metals, $\mathrm{Pr} \approx 0.01$; for high-viscosity oils,
$\mathrm{Pr} \approx 1000$. Prandtl is a fluid property, not a
flow property; it shapes the ratio of velocity boundary-layer
thickness to thermal boundary-layer thickness.

\engterm{Nusselt number} $\mathrm{Nu} = h L/k$ (convective heat
transfer to conductive). For free convection from a vertical
plate, $\mathrm{Nu} = C(\mathrm{Gr}\, \mathrm{Pr})^{n}$ with $C
\sim 0.59$, $n = 0.25$ in laminar (Churchill-Chu correlation);
for forced convection in pipe flow at the Dittus-Boelter
correlation, $\mathrm{Nu} = 0.023\,\mathrm{Re}^{0.8}\,
\mathrm{Pr}^{0.4}$ for heating.

\engterm{Grashof number} $\mathrm{Gr} = g\beta(T_{s} - T_{\infty})
L^{3}/\nu^{2}$ (buoyancy to viscous forces in natural convection).
The product $\mathrm{Ra} = \mathrm{Gr}\,\mathrm{Pr}$ (Rayleigh
number) governs the transition from conduction-dominated to
convection-dominated transport in horizontal layers
($\mathrm{Ra} > 1708$ for B\'enard convection onset) and from
laminar to turbulent natural convection on vertical plates
($\mathrm{Ra} > 10^{9}$).

\engterm{P\'eclet number} $\mathrm{Pe} = vL/D$ (advection to
diffusion). $\mathrm{Pe} = \mathrm{Re}\,\mathrm{Pr}$ for thermal
transport; $\mathrm{Pe} = \mathrm{Re}\,\mathrm{Sc}$ for mass
transport. Determines whether mixing in a pipe is diffusion-
limited ($\mathrm{Pe} < 1$) or advection-limited ($\mathrm{Pe}
\gg 1$).

\engterm{Schmidt number} $\mathrm{Sc} = \nu/D$ (momentum to mass
diffusivity). The mass-transfer analogue of $\mathrm{Pr}$. For
gases, $\mathrm{Sc} \approx 1$; for liquids, $\mathrm{Sc} \approx
10^{3}$. Volume V chapter $9$ develops the working applications.

\engterm{Sherwood number} $\mathrm{Sh} = k_{m} L/D$ (convective
mass transfer to diffusive). The mass-transfer analogue of
$\mathrm{Nu}$. Working correlations follow the same form:
$\mathrm{Sh} = C\,\mathrm{Re}^{m}\,\mathrm{Sc}^{n}$ with
correlation-dependent coefficients.

\engterm{Knudsen number} $\mathrm{Kn} = \lambda/L$ (molecular mean
free path to characteristic length). $\mathrm{Kn} < 10^{-3}$:
continuum regime, Navier-Stokes applies. $\mathrm{Kn} > 10$:
free-molecular regime, kinetic theory required. Determines the
applicability of continuum fluid mechanics for MEMS, vacuum
systems, and aerospace high-altitude flows.

\engterm{E\"otv\"os (Bond) number} $\mathrm{Eo} = \rho g L^{2}/
\sigma$ (gravity to surface tension on a length scale). At
$\mathrm{Eo} \ll 1$, surface tension dominates (small drops are
spherical); at $\mathrm{Eo} \gg 1$, gravity dominates (large
puddles are flat).

\engterm{Stokes number} $\mathrm{Stk} = \tau_{p} v/L$ (particle
relaxation time times velocity over length scale, with $\tau_{p}
= \rho_{p} d_{p}^{2}/(18\mu)$). $\mathrm{Stk} \ll 1$: particles
follow streamlines (aerosol). $\mathrm{Stk} \gg 1$: particles
ignore streamlines (impactor in particle sampling).

The selection of which named groups appear in a given Buckingham-
pi analysis is the engineering judgement on which physical
mechanisms matter for the problem. A reader who recognises the
groups as they emerge from the dimensional analysis can name
the regime, locate the working correlation in the relevant
handbook, and predict the order of magnitude of the answer before
doing any calculation. The discipline that scales from a problem
statement to a working answer is the same as the discipline that
scales from a measurement to a usable engineering quantity:
identify the variables, identify the regime, identify the
correlation, identify the uncertainty.

\subsection{Reading correlations and the Moody diagram}
\pathtag{standard}

A working engineer reads dimensionless-group correlations the way a
working mechanic reads a torque-spec table: by regime, by exponent,
and by validity bound. We walk the three reading skills with a
$\mathrm{Nu} = C \cdot \mathrm{Ra}^{a} \cdot \mathrm{Pr}^{b}$
correlation in hand and then trace the same disciplines through the
Moody diagram for pipe-flow friction factor, which is the most
heavily-used dimensionless-group chart in mechanical engineering.

\paragraph{What the exponents tell you.} A correlation written as
$\mathrm{Nu} = C \cdot \mathrm{Ra}^{a} \cdot \mathrm{Pr}^{b}$
encodes how the heat-transfer coefficient $h$ scales with the
dimensional variables. From $\mathrm{Nu} = h L / k$ and
$\mathrm{Ra} = g \beta \Delta T L^{3} / (\nu \alpha)$, the
correlation rearranges to
\[
h = C \cdot \frac{k}{L} \cdot
   \left(\frac{g \beta \Delta T L^{3}}{\nu \alpha}\right)^{a}
   \cdot \mathrm{Pr}^{b}
= C \cdot k \cdot
  \frac{(g \beta \Delta T)^{a}}{\nu^{a} \alpha^{a}}
  \cdot L^{3a - 1} \cdot \mathrm{Pr}^{b}.
\]
The reader extracts three working scaling laws from the exponent
$a$ alone. First, the dependence on the length scale $L$ is
$L^{3a - 1}$. For laminar free convection on a vertical plate the
Churchill-Chu result gives $a = 1/4$, so $h \sim L^{-1/4}$: the
heat-transfer coefficient \emph{decreases} with plate height. For
turbulent free convection $a = 1/3$, so $h \sim L^{0}$ is
independent of plate height (the famous height-independent natural-
convection regime). For forced-convection pipe flow at Dittus-
Boelter the exponent on $\mathrm{Re}$ is $0.8$, giving $h \sim
v^{0.8} D^{-0.2}$: doubling the flow velocity raises $h$ by
$2^{0.8} \approx 1.74$. Second, the dependence on driving
temperature difference is $(\Delta T)^{a}$. For laminar free
convection $h$ grows as $\Delta T^{1/4}$: the heat-transfer
coefficient is only weakly dependent on the temperature difference,
even though the heat flux $q'' = h \Delta T$ scales as
$\Delta T^{5/4}$. Third, the exponent $b$ on the Prandtl number
distinguishes a momentum-controlled boundary layer
($b > 0$) from a thermally controlled one ($b < 0$); for
Dittus-Boelter, $b = 0.4$ for heating and $b = 0.3$ for cooling,
the asymmetry reflecting the temperature dependence of the
boundary-layer thicknesses.

\paragraph{What the prefactor and the regime envelope tell you.}
The prefactor $C$ is the bridge between dimensional structure and
empirical reality. A reader who finds $C \approx 0.59$ for laminar
free convection on a vertical plate and $C \approx 0.10$ for the
same geometry in the turbulent regime sees the order-of-magnitude
shift across regimes and knows to plot the correlation on log-log
axes before extrapolating. Every published correlation carries a
validity envelope: Churchill-Chu specifies $\mathrm{Ra} \leq 10^{9}$,
Dittus-Boelter specifies $\mathrm{Re} > 10^{4}$ and $0.6 <
\mathrm{Pr} < 160$ and $L/D > 60$. The envelope is the warranty.
Outside it, the correlation has neither calibration nor uncertainty;
a value computed there is a guess.

\paragraph{The Moody diagram annotated.}
\Cref{fig:vol01:ch02:moody} shows the friction factor $f$ for pipe
flow as a function of Reynolds number and relative roughness
$e/D$. The diagram is the canonical engineering reference for
pressure drop in single-phase incompressible pipe flow
\cite{paper:moody-1944}. Three regimes appear.

The laminar regime below $\mathrm{Re} \approx 2300$ obeys the
Hagen-Poiseuille relation $f = 64/\mathrm{Re}$, a straight line of
slope $-1$ on log-log axes. The relation is exact for fully-
developed laminar flow in a smooth round pipe; relative roughness
plays no role because the laminar viscous sublayer covers the wall
asperities completely.

The transition band $2300 < \mathrm{Re} < 4000$ is the unstable
zone in which the flow may be laminar or turbulent depending on
upstream disturbance, pipe roughness, and inlet conditions. A
designer does not operate in this band; pumps and piping
diameters are chosen to fall on one side or the other.

Above $\mathrm{Re} \approx 4000$, turbulence sets in and $f$
follows the Colebrook equation
\cite{paper:colebrook-1939}:
\[
\frac{1}{\sqrt{f}} = -2 \log_{10}\!\left(\frac{e/D}{3.7}
  + \frac{2.51}{\mathrm{Re} \sqrt{f}}\right).
\]
At low Reynolds number the second term inside the logarithm
dominates and $f$ depends on $\mathrm{Re}$; this is the
smooth-turbulent regime where Blasius' explicit
$f = 0.316 \,\mathrm{Re}^{-1/4}$ applies for $4000 < \mathrm{Re} <
10^{5}$. At high Reynolds number the first term dominates and the
$\mathrm{Re}$ dependence disappears; this is the fully rough
regime where $f$ is set by $e/D$ alone. Between the two, both terms
contribute; the family of curves on the Moody diagram traces this
transition for each value of $e/D$.

\paragraph{A worked friction-factor extraction.} A typical
engineering problem: a $6 \,\text{in}$ commercial steel pipe
(schedule 40, internal diameter $D = 154 \,\si{\milli\meter}$) at
$\mathrm{Re} = 10^{6}$. The Crane Technical Paper 410 value for
commercial steel roughness is $e \approx 46 \,\si{\micro\meter}$,
giving relative roughness
$e/D = 4.6 \times 10^{-5}/0.154 = 3.0 \times 10^{-4}$.
The Colebrook equation is implicit in $f$; the explicit
Swamee-Jain approximation \cite{paper:swamee-jain-1976},
\[
f = \frac{0.25}{\left[\log_{10}\!\left(\dfrac{e/D}{3.7}
  + \dfrac{5.74}{\mathrm{Re}^{0.9}}\right)\right]^{2}},
\]
gives the engineering answer in one evaluation. The inner term:
$(e/D)/3.7 = 3.0 \times 10^{-4}/3.7 = 8.11 \times 10^{-5}$;
$5.74/\mathrm{Re}^{0.9} = 5.74/(10^{6})^{0.9}
= 5.74/2.51 \times 10^{5} = 2.29 \times 10^{-5}$.
Sum: $1.04 \times 10^{-4}$. Logarithm: $\log_{10}(1.04 \times 10^{-4})
= -3.98$. Squared: $15.85$. Friction factor:
$f = 0.25/15.85 = 0.0158$, rounding to $f \approx 0.016$. The
Moody-diagram reading at the same point is $f \approx 0.014$ to
$0.016$, consistent within reading uncertainty. Either answer is
useful; both are inside the Colebrook uncertainty envelope.

\paragraph{A worked pressure-drop calculation.} A $100 \,\si
{\meter}$ horizontal run of the same pipe at $v = 2 \,\si{\meter
\per\second}$ carrying water at $20 \,\degC$ ($\rho = 998 \,\si
{\kilogram\per\meter\cubed}$, $\mu = 1.0 \times 10^{-3} \,\si
{\pascal\second}$). The Reynolds number is
$\mathrm{Re} = \rho v D / \mu = 998 \times 2 \times 0.154/10^{-3}
= 3.07 \times 10^{5}$, comfortably turbulent. The Darcy-Weisbach
pressure drop is
\[
\Delta p = f \cdot \frac{L}{D} \cdot \frac{\rho v^{2}}{2}
= 0.016 \times \frac{100}{0.154} \times \frac{998 \times 2^{2}}{2}
= 0.016 \times 649 \times 1996 \approx 2.07 \times 10^{4} \,\si
{\pascal} \approx 20.7 \,\si{\kilo\pascal}.
\]
For a pump delivering this flow against this drop, the hydraulic
power is $P_{h} = Q \cdot \Delta p$, with
$Q = v A = 2 \times \pi (0.154)^{2}/4 = 0.0373 \,\si{\meter
\cubed\per\second}$, giving
$P_{h} = 0.0373 \times 2.07 \times 10^{4} = 771 \,\si{\watt}
\approx 0.8 \,\si{\kilo\watt}$. The pump's shaft power is
$P_{s} = P_{h}/\eta$ with pump efficiency $\eta \approx 0.7$ for a
typical centrifugal pump in this duty, so $P_{s} \approx 1.1 \,\si
{\kilo\watt}$. The engineer who carries the dimensional check
through to the wattage can size the motor.

\paragraph{The diagram as a Buckingham-pi result.} The Moody chart
is the empirical $f = \Phi(\mathrm{Re}, e/D)$ relationship that
dimensional analysis predicted, plotted on log-log axes. The
variable list for the original problem was
$\{\Delta p, L, D, v, \rho, \mu, e\}$: seven variables, three base
dimensions, four dimensionless groups
($f, \mathrm{Re}, L/D, e/D$). The $L/D$ dependence is folded into
the Darcy-Weisbach formula by carrying $L/D$ outside the friction
factor. The remaining $f(\mathrm{Re}, e/D)$ is the chart. Every
pipe-flow pressure-drop calculation in the next eight decades
of engineering practice has been a single chart-reading and a
multiplication. The economy of dimensional analysis, measured in
hours of engineer time saved per design, is in the millions.

\input{volumes/01-quantity/ch02-units-and-dimensions/figures/fig-moody-diagram.tex}

\Cref{fig:vol01:ch02:flow-regimes} summarises the regime map in
$\mathrm{Re}$ and $\mathrm{Ma}$ that selects which dimensionless
groups the engineer must track for a given problem. A reader who
locates the design point on this map before opening a handbook
saves time looking in the wrong section.

\input{volumes/01-quantity/ch02-units-and-dimensions/figures/fig-flow-regimes.tex}

\subsection{Worked example: the simple pendulum}

A pendulum of length $L$ swings under gravity $g$. The bob has mass
$m$. We want the period $T$.

\begin{itemize}[leftmargin=*]
  \item Variables: $T, L, g, m$. So $n = 4$.
  \item Dimensions: $[T] = [T]$; $[L] = [L]$; $[g] = [L][T]^{-2}$;
    $[m] = [M]$. Three independent dimensions appear: $[L], [T],
    [M]$. So $k = 3$, and $n - k = 1$.
  \item One dimensionless group. Choose $L, g, m$ as repeating
    variables; they span all three dimensions independently.
  \item Form $\pi_{1}$ from $T$ and the repeating variables:
    $\pi_{1} = T \cdot L^{a} \cdot g^{b} \cdot m^{c}$. Set the
    dimensions of $\pi_{1}$ to zero in each base dimension:
    \[
    [T]: \quad 1 + (-2)b = 0, \qquad
    [L]: \quad a + b = 0, \qquad
    [M]: \quad c = 0.
    \]
    Solve: $b = 1/2$, $a = -1/2$, $c = 0$. So
    \[
    \pi_{1} = T \sqrt{\frac{g}{L}}.
    \]
\end{itemize}

The conclusion: $T \sqrt{g/L}$ is constant, or equivalently
$T \propto \sqrt{L/g}$. The mass drops out, which is correct: the
pendulum's period is independent of the bob's mass. The constant
$2\pi$ that appears in the small-amplitude solution is exactly the
prefactor dimensional analysis cannot give us. The prefactor would
also depend on amplitude for large swings; for the pendulum the
prefactor is $2\pi$ in the small-angle limit and grows slowly with
amplitude.

\subsection{Worked example: drag on a sphere}

A sphere of diameter $D$ moves through a fluid of density $\rho$ and
dynamic viscosity $\mu$ at speed $v$. We want the drag force $F$.

\begin{itemize}[leftmargin=*]
  \item Variables: $F, \rho, v, \mu, D$. So $n = 5$.
  \item Dimensions: $[F] = [M][L][T]^{-2}$; $[\rho] = [M][L]^{-3}$;
    $[v] = [L][T]^{-1}$; $[\mu] = [M][L]^{-1}[T]^{-1}$; $[D] = [L]$.
    Three independent dimensions appear: $[M], [L], [T]$. So $k =
    3$, and $n - k = 2$.
  \item Two dimensionless groups. Choose $\rho, v, D$ as repeating
    variables.
  \item First group, with $F$:
    \[
    \pi_{1} = F \cdot \rho^{a} v^{b} D^{c}.
    \]
    Setting dimensions to zero:
    \[
    [M]: \quad 1 + a = 0, \qquad
    [L]: \quad 1 - 3a + b + c = 0, \qquad
    [T]: \quad -2 - b = 0.
    \]
    Solve: $a = -1$, $b = -2$, $c = -2$. So
    \[
    \pi_{1} = \frac{F}{\rho v^{2} D^{2}}.
    \]
    This is the drag coefficient up to a factor of $\pi/8$ (because
    one usually defines $C_{D} = F / (\tfrac{1}{2} \rho v^{2} A)$
    with $A = \pi D^{2}/4$).
  \item Second group, with $\mu$:
    \[
    \pi_{2} = \mu \cdot \rho^{a} v^{b} D^{c}.
    \]
    Solving as before: $a = -1$, $b = -1$, $c = -1$. So
    \[
    \pi_{2} = \frac{\mu}{\rho v D},
    \]
    which is the inverse Reynolds number. By convention, $\mathrm{Re}
    = \rho v D / \mu$ is used instead of $\pi_{2}$, since both are
    dimensionless.
\end{itemize}

The conclusion: the drag coefficient is a function of Reynolds number
alone:

\[
C_{D} = \Phi(\mathrm{Re}).
\]

Dimensional analysis gives the form, not the curve. Experiments
measure $\Phi$ and also test whether the variable list was complete:
roughness, compressibility, wall effects, and unsteady motion can all
add dimensionless groups. We return to the measured curve in chapter
12 of Volume III. \Cref{fig:vol01:ch02:dim-matrix} shows the
dimensional matrix for this problem; \Cref{fig:vol01:ch02:drag-curve}
shows the measured drag-coefficient curve spanning seven orders of
magnitude in Reynolds number.

\input{volumes/01-quantity/ch02-units-and-dimensions/figures/fig-dimensional-matrix.tex}
\input{volumes/01-quantity/ch02-units-and-dimensions/figures/fig-drag-curve.tex}

\subsection{Worked example: free convection, the Nusselt-Rayleigh
relation}\pathtag{standard}

The pendulum and drag-on-a-sphere examples involve at most five
variables and recover famous dimensionless groups. Heat transfer in
free convection demands a longer variable list, produces three
dimensionless groups, and shows how dimensional analysis collapses a
six-variable experimental space onto a two-variable correlation curve
that fits in a single Moody-style chart. We work the case at the
detail an engineer would carry into the laboratory or the design
review.

\paragraph{Physical setting.} A hot vertical plate is immersed in a
still ambient fluid. The plate has height $L$ and surface temperature
$T_{s}$; the ambient fluid is at temperature $T_{\infty}$. The
temperature difference $\Delta T = T_{s} - T_{\infty}$ drives a
buoyant flow that carries heat away from the plate. We want the
heat-transfer coefficient $h$ between the plate and the fluid, or
equivalently the heat-transfer rate per area $q'' = h \Delta T$.

\paragraph{Variable list.} The physical mechanism involves buoyancy
(driven by $\Delta T$ and the fluid's thermal-expansion coefficient
$\beta$), gravity $g$ (which converts the buoyancy into a body
force), thermal conduction (set by the fluid's thermal conductivity
$k$), thermal capacity (set by the fluid's heat capacity $c_{p}$ and
density $\rho$, or together the volumetric heat capacity $\rho c_{p}$),
and the resistance to flow (set by the dynamic viscosity $\mu$). The
length scale of the problem is $L$. The variable list is therefore
\[
h, \quad L, \quad k, \quad \mu, \quad \rho, \quad c_{p}, \quad
g \beta \Delta T.
\]
The last variable groups gravity, the thermal-expansion coefficient,
and the temperature difference into a single buoyancy-acceleration
term; the engineering convention is to use this combined variable
because it is the physically meaningful driving acceleration of the
free convection. Seven variables in total.

\paragraph{Dimensional signatures.} Writing each variable's dimensions
in $[M], [L], [T], [\Theta]$ (where $[\Theta]$ is temperature):
\begin{align*}
[h] &= [M][T]^{-3}[\Theta]^{-1}
\quad (\text{heat flux per temperature difference}), \\
[L] &= [L], \\
[k] &= [M][L][T]^{-3}[\Theta]^{-1}
\quad (\text{thermal conductivity, } W/(m\cdot K)), \\
[\mu] &= [M][L]^{-1}[T]^{-1}, \\
[\rho] &= [M][L]^{-3}, \\
[c_{p}] &= [L]^{2}[T]^{-2}[\Theta]^{-1}
\quad (\text{specific heat, } J/(kg\cdot K)), \\
[g \beta \Delta T] &= [L][T]^{-2}.
\end{align*}
Four base dimensions appear and seven variables are present; the
Buckingham pi theorem predicts $7 - 4 = 3$ dimensionless groups.

\paragraph{Repeating-variable choice.} The four repeating variables
must span the four base dimensions independently. A working choice is
$L, \mu, k, \rho$ (the length scale, two transport properties, and
the density). The remaining variables $h, c_{p}, g\beta\Delta T$ each
become one of the dimensionless groups.

\paragraph{Group $\pi_{1}$ (Nusselt number).} Form $\pi_{1} = h \cdot
L^{a} \mu^{b} k^{c} \rho^{d}$. Setting each base-dimension exponent
to zero,
\[
[M]: 1 + b + c + d = 0, \quad
[L]: a - b + c - 3d = 0, \quad
[T]: -3 - b - 3c = 0, \quad
[\Theta]: -1 - c = 0.
\]
The $[\Theta]$ equation gives $c = -1$. The $[T]$ equation gives
$b = 0$. The $[M]$ equation gives $d = 0$. The $[L]$ equation gives
$a = 1$. Therefore $\pi_{1} = h L / k$, which is the \engterm{Nusselt
number} $\mathrm{Nu}$: the ratio of convective to conductive heat
transfer at the surface.

\paragraph{Group $\pi_{2}$ (Prandtl number).} Form $\pi_{2} = c_{p}
\cdot L^{a} \mu^{b} k^{c} \rho^{d}$. The system is
\[
[M]: b + c + d = 0, \quad
[L]: 2 + a - b + c - 3d = 0, \quad
[T]: -2 - b - 3c = 0, \quad
[\Theta]: -1 - c = 0.
\]
The $[\Theta]$ equation gives $c = -1$. The $[T]$ equation gives
$b = 1$. The $[M]$ equation gives $d = 0$. The $[L]$ equation gives
$a = 0$. Therefore $\pi_{2} = c_{p} \mu / k$, which is the
\engterm{Prandtl number} $\mathrm{Pr}$: the ratio of momentum
diffusivity to thermal diffusivity.

\paragraph{Group $\pi_{3}$ (Grashof number).} Form $\pi_{3} = (g\beta
\Delta T) \cdot L^{a} \mu^{b} k^{c} \rho^{d}$. The system is
\[
[M]: b + c + d = 0, \quad
[L]: 1 + a - b + c - 3d = 0, \quad
[T]: -2 - b - 3c = 0, \quad
[\Theta]: -c = 0.
\]
The $[\Theta]$ equation gives $c = 0$. The $[T]$ equation gives
$b = -2$. The $[M]$ equation gives $d = 2$. The $[L]$ equation gives
$a = 3$. Therefore $\pi_{3} = g \beta \Delta T \cdot L^{3} \rho^{2}
/ \mu^{2}$, which is the \engterm{Grashof number} $\mathrm{Gr}$: the
ratio of buoyant to viscous forces in the free-convection boundary
layer.

\paragraph{The Rayleigh number.} The product $\mathrm{Ra} =
\mathrm{Gr} \cdot \mathrm{Pr} = g \beta \Delta T L^{3} \rho c_{p} /
(k \mu) = g \beta \Delta T L^{3} / (\nu \alpha)$, where $\nu = \mu/
\rho$ is the kinematic viscosity and $\alpha = k/(\rho c_{p})$ is the
thermal diffusivity, is the \engterm{Rayleigh number}. The convention
in free convection is to use $\mathrm{Ra}$ rather than $\mathrm{Gr}$
because the empirical correlations of the Nusselt number against
$\mathrm{Ra}$ are nearly Prandtl-number-independent for a wide range
of common fluids.

\paragraph{The collapse.} The dimensional argument concludes that
\[
\mathrm{Nu} = \Phi(\mathrm{Ra}, \mathrm{Pr})
\]
for some unknown function $\Phi$. The seven-variable experimental
space has collapsed to a three-variable function and (for the
Prandtl-weak case) effectively to a two-variable curve
$\mathrm{Nu}(\mathrm{Ra})$ that can be plotted on log-log axes as a
single line. This is the Moody-diagram analogue for free convection:
one curve summarises every free-convection experiment on a vertical
plate for every fluid, every plate height, every temperature
difference, within the validity envelope.

\paragraph{The published correlation.} For laminar free convection
on a vertical isothermal plate, Churchill and Chu's correlation gives
\[
\mathrm{Nu}_{L}^{1/2}
= 0.68 + \frac{0.670 \, \mathrm{Ra}_{L}^{1/4}}
{[1 + (0.492/\mathrm{Pr})^{9/16}]^{4/9}}
\quad \text{for } \mathrm{Ra}_{L} \leq 10^{9}.
\]
The form is empirical: the prefactors $0.68$, $0.670$, $0.492$, and
the exponents $9/16$, $4/9$ are fit to experimental data. The
dominant scaling is $\mathrm{Nu} \sim \mathrm{Ra}^{1/4}$ at large
Rayleigh number, which dimensional analysis cannot derive (it
emerges from the boundary-layer-equation similarity solution).
Tabulated values and an extensive discussion appear in Incropera
and DeWitt \cite{text:incropera2007}.

\paragraph{Worked design case.} Estimate the heat-transfer
coefficient from a $L = 0.5 \,\si{\meter}$-tall vertical aluminium
plate at $T_{s} = 80 \,\si{\degreeCelsius}$ in still air at
$T_{\infty} = 20 \,\si{\degreeCelsius}$.

Air properties at the mean film temperature $T_{f} = (80 + 20)/2 =
50 \,\si{\degreeCelsius}$: $\nu = 1.79 \times 10^{-5} \,\si{\meter
\squared\per\second}$, $\alpha = 2.54 \times 10^{-5} \,\si{\meter
\squared\per\second}$, $\beta = 1/T_{f} = 1/323 \,\si{\per\kelvin}
= 3.10 \times 10^{-3} \,\si{\per\kelvin}$, $k = 0.0282
\,\si{\watt\per\meter\per\kelvin}$, $\mathrm{Pr} = \nu/\alpha = 0.70$.
The Rayleigh number is
$\mathrm{Ra}_{L} = g \beta \Delta T L^{3} / (\nu \alpha) =
9.81 \times 3.10 \times 10^{-3} \times 60 \times 0.5^{3} / (1.79
\times 10^{-5} \times 2.54 \times 10^{-5}) = 5.0 \times 10^{8}$.
This is within the laminar regime ($\mathrm{Ra} \leq 10^{9}$), so
Churchill-Chu applies.

The Churchill-Chu correlation gives
$\mathrm{Nu}_{L}^{1/2} = 0.68 + 0.670 \times (5.0 \times 10^{8})^{1/4}
/ [1 + (0.492/0.70)^{9/16}]^{4/9}$.
The fourth root of the Rayleigh number is
$(5.0 \times 10^{8})^{1/4} \approx 150$. The denominator's bracketed
factor evaluates to approximately $1.49$ raised to $4/9 \approx 1.20$.
Then $\mathrm{Nu}_{L}^{1/2} \approx 0.68 + 0.670 \times 150 / 1.20
\approx 0.68 + 83.75 \approx 84.4$, giving $\mathrm{Nu}_{L} \approx
7124$. The heat-transfer coefficient is
$h = \mathrm{Nu}_{L} \cdot k / L = 7124 \times 0.0282 / 0.5
\approx 402 \,\si{\watt\per\meter\squared\per\kelvin}$.

The heat flux from the plate is $q'' = h \Delta T = 402 \times 60
\approx 24{,}100 \,\si{\watt\per\meter\squared}$. The total
heat-transfer rate, if the plate has area $A = L \times W$ with
some width $W$, is $Q = q'' \times A$. For $W = 0.5 \,\si{\meter}$
(a square plate), $Q \approx 6 \,\si{\kilo\watt}$.

\paragraph{Sanity check by limits.} The Nusselt number for pure
conduction (no flow) is of order one; the result $\mathrm{Nu}_{L}
\approx 7000$ says free convection dominates conduction by three
orders of magnitude. The heat-transfer coefficient
$h \approx 400 \,\si{\watt\per\meter\squared\per\kelvin}$ is at the
high end for natural convection in air; published tables for
free-convection heat transfer in air range from approximately
$5$ to $25 \,\si{\watt\per\meter\squared\per\kelvin}$ for low-
temperature surfaces with smaller $\Delta T$. The factor-of-twenty
elevation in $h$ over the textbook range reflects the relatively
large $L = 0.5 \,\si{\meter}$ and the relatively large $\Delta T =
60 \,\si{\kelvin}$ of the present case. The Rayleigh-number scaling
$\mathrm{Nu} \sim \mathrm{Ra}^{1/4}$ implies $h \sim L^{-1/4} \cdot
(g \beta \Delta T)^{1/4}$ on the high-Rayleigh limit, so doubling
$\Delta T$ or doubling $L$ raises $h$ by the same factor of
approximately $1.19$. The scaling is verifiable by repeating the
correlation at $\Delta T = 30 \,\si{\kelvin}$ and confirming the
result is closer to $h \approx 340 \,\si{\watt\per\meter\squared
\per\kelvin}$.

\paragraph{What dimensional analysis delivered.} The seven-variable
free-convection problem collapsed to three dimensionless groups
($\mathrm{Nu}$, $\mathrm{Pr}$, $\mathrm{Gr}$ or equivalently
$\mathrm{Nu}$, $\mathrm{Pr}$, $\mathrm{Ra}$), and the engineering
correlation gives the Nusselt number as a function of two of them.
A single curve covers every laminar free-convection experiment on a
vertical plate, for every fluid, every plate height, every
temperature difference within the laminar envelope. The economy of
experimentation is the same economy that turns drag-curve
experiments on spheres into the single $C_{D}(\mathrm{Re})$ chart
in Volume III. The dimensional argument is the framework that makes
the empirical correlation a universal engineering result rather than
a one-off measurement.

\subsection{Worked example: pump affinity laws}
\pathtag{standard}

A centrifugal pump's head $H$, flow $Q$, shaft power $P$, and the
trio of speed $\omega$, impeller diameter $D$, and fluid properties
($\rho$, $\mu$, $g$) form a closed dimensional system. Three
dimensionless groups exhaust it; the resulting relationships, called
the \engterm{affinity laws}, sit behind every pump selection,
every variable-speed-drive retrofit, and every impeller-trim decision
in industrial fluid handling.

\paragraph{Physical setting.} A centrifugal pump moves a fluid of
density $\rho$ and viscosity $\mu$ against gravity $g$. The impeller
rotates at angular speed $\omega$ and has characteristic diameter
$D$. The pump produces a head $H$ (the rise in mechanical-energy
height per unit weight, $\si{\meter}$ of column), passes a volumetric
flow $Q$, and absorbs shaft power $P$. The aim is to predict how
$H$, $Q$, and $P$ scale when the operator changes $\omega$ or $D$.

\paragraph{Variable list.} Eight variables:
\[
H, Q, P, \omega, D, \rho, \mu, g.
\]

\paragraph{Dimensional signatures.}
\begin{align*}
[H] &= [L], & [Q] &= [L]^{3}[T]^{-1}, & [P] &= [M][L]^{2}[T]^{-3}, \\
[\omega] &= [T]^{-1}, & [D] &= [L], & [\rho] &= [M][L]^{-3}, \\
[\mu] &= [M][L]^{-1}[T]^{-1}, & [g] &= [L][T]^{-2}.
\end{align*}
Three base dimensions appear ($[M], [L], [T]$); eight variables;
$8 - 3 = 5$ dimensionless groups.

\paragraph{Repeating-variable choice.} Choose $\rho$, $\omega$, $D$.
These span $[M]$, $[T]$, and $[L]$ independently. The remaining five
variables ($H, Q, P, \mu, g$) each give one dimensionless group.

\paragraph{The five groups.} Solving the dimensional system for each
non-repeating variable:
\begin{align*}
\pi_{1} &= \frac{H}{D} && \text{(head coefficient, geometric)}, \\
\pi_{2} &= \frac{Q}{\omega D^{3}} && \text{(flow coefficient)},
  \quad \phi \\
\pi_{3} &= \frac{P}{\rho \omega^{3} D^{5}}
  && \text{(power coefficient)}, \quad \chi \\
\pi_{4} &= \frac{\rho \omega D^{2}}{\mu}
  && \text{(rotational Reynolds number)}, \quad \mathrm{Re}_{\omega} \\
\pi_{5} &= \frac{g H}{\omega^{2} D^{2}}
  && \text{(head-rise coefficient)}, \quad \psi.
\end{align*}
The convention is to use the head-rise coefficient $\psi = gH/
(\omega^{2} D^{2})$ rather than $H/D$, because $\psi$ carries the
mechanical-energy meaning directly. Replacing $\pi_{1}$ with
$\psi$ leaves five independent groups.

\paragraph{The structural relationship.} For dimensionally similar
operating points,
\[
\psi = \Phi(\phi, \mathrm{Re}_{\omega}), \quad
\chi = \Psi(\phi, \mathrm{Re}_{\omega}),
\]
with two unknown functions $\Phi$ and $\Psi$. At high
$\mathrm{Re}_{\omega}$, where viscous effects are subdominant,
the Reynolds dependence drops and
$\psi = \Phi(\phi)$, $\chi = \Psi(\phi)$: the dimensionless head
and dimensionless power depend only on the dimensionless flow. The
pump characteristic curves published by the manufacturer
($H$ versus $Q$ at one speed, with efficiency contours overlaid)
are the dimensional forms of $\Phi$ and $\Psi$ at a chosen
$\omega$ and $D$.

\paragraph{The affinity laws.} Hold $\phi$ constant by adjusting
$\omega$ and $D$ in step. Then $\psi$ and $\chi$ are constant.
Reading off the scaling of the dimensional variables,
\[
\boxed{\frac{H_{2}}{H_{1}} = \left(\frac{\omega_{2}}{\omega_{1}}\right)^{2}
  \left(\frac{D_{2}}{D_{1}}\right)^{2}, \quad
\frac{Q_{2}}{Q_{1}} = \frac{\omega_{2}}{\omega_{1}}
  \left(\frac{D_{2}}{D_{1}}\right)^{3}, \quad
\frac{P_{2}}{P_{1}} = \frac{\rho_{2}}{\rho_{1}}
  \left(\frac{\omega_{2}}{\omega_{1}}\right)^{3}
  \left(\frac{D_{2}}{D_{1}}\right)^{5}.}
\]
The affinity laws follow directly from the constancy of $\psi$,
$\phi$, and $\chi$ at the dimensionally similar operating point.
Karassik gives the standard derivation and a wealth of working
examples \cite{text:ch02t3-karassik-pumps}.

\paragraph{Deviation at low Reynolds.} The Reynolds-dependence of
$\Phi$ and $\Psi$ becomes audible below $\mathrm{Re}_{\omega} \sim
10^{4}$, when viscous losses in the volute and impeller passages
shift the position of the best-efficiency point and reduce the
attainable head coefficient. For viscous fluids such as heavy
crude oils ($\mu = 10$ to $1000$ times water), the Hydraulic
Institute publishes correction factors for $H$, $Q$, and $\eta$
that transform the water-tested characteristic curves to viscous
service. The affinity laws as written above presume
$\mathrm{Re}_{\omega}$ large enough that the viscous correction
is negligible.

\paragraph{Worked numerical example.} A Goulds 3196 process pump,
6$\times$8-15 frame (suction $8\,\text{in}$, discharge $6\,\text{in}$,
impeller diameter $D_{1} = 15\,\text{in} = 0.381 \,\si{\meter}$),
operates at the duty point $\omega_{1} = 1750\,\text{rpm}$,
$Q_{1} = 0.063 \,\si{\meter\cubed\per\second}$ (about
$1000 \,\si{US gallon\per\minute}$), $H_{1} = 60 \,\si{\meter}$,
shaft power $P_{1} = 50 \,\si{\kilo\watt}$, on water
\cite{text:ch02t3-goulds-3196}.

The plant retrofits a variable-frequency drive and runs the same
pump at $\omega_{2} = 1450 \,\text{rpm}$ (a reduction to about
$83\%$ of original speed) to reduce the flow to the new process
requirement.

By the affinity laws with $D_{2} = D_{1}$:
\begin{align*}
\frac{Q_{2}}{Q_{1}} &= \frac{1450}{1750} = 0.829, \quad
Q_{2} = 0.052 \,\si{\meter\cubed\per\second}, \\
\frac{H_{2}}{H_{1}} &= 0.829^{2} = 0.687, \quad
H_{2} = 41.2 \,\si{\meter}, \\
\frac{P_{2}}{P_{1}} &= 0.829^{3} = 0.569, \quad
P_{2} = 28.5 \,\si{\kilo\watt}.
\end{align*}
The cubic scaling of power with speed is the engineering bottom
line behind the variable-frequency-drive retrofit case: at $83\%$
speed the pump draws $57\%$ of the original power, even though it
moves $83\%$ of the original flow. The hydraulic efficiency
$\eta = \rho g Q H / P$ remains constant under the affinity laws
(check: $\rho g Q H / P \propto (1)(0.829)(0.687)/(0.569) = 1.00$),
which is why the operating point at the new speed lies on a
constant-efficiency line of the pump characteristic chart.

\paragraph{The dimensional argument's economy.} A pump manufacturer
publishes the characteristic curve at one rated speed and one
nominal impeller diameter. The Buckingham-pi result lets the
engineer transport the curve to any other speed and any
geometrically similar impeller diameter, within the validity of
the high-Reynolds approximation. The economy is the same one
that the Moody diagram secured for pipe-flow pressure drop and that
the Churchill-Chu correlation secured for free convection: one
chart, every operating point.

\subsection{Worked example: wind-tunnel scaling for an airfoil}
\pathtag{standard}

A scale-model wing in a wind tunnel produces lift and drag
coefficients that match the full-scale prototype only if the
governing dimensionless groups match. For a NACA 0012 airfoil at
the design point of a small unmanned aircraft, we walk the matching
discipline and the practical compromises that the impossible-to-
match cases force.

\paragraph{Physical setting.} A wing of chord $c$, thickness ratio
$t/c$, span $b$, and angle of attack $\alpha$ moves through air at
density $\rho$, viscosity $\mu$, sound speed $a$, and freestream
velocity $V$. The output of interest is the lift coefficient
$C_{L} = L/(\tfrac{1}{2} \rho V^{2} c b)$ and the drag coefficient
$C_{D} = D/(\tfrac{1}{2} \rho V^{2} c b)$. The lift and drag are
single-valued functions of the dimensionless geometric and flow
parameters. Anderson's \emph{Fundamentals of Aerodynamics} gives
the working development \cite{text:ch02t3-anderson-fundamentals}.

\paragraph{Variable list and dimensionless groups.} The relevant
variables for fixed geometry are $V, c, \rho, \mu, a$. Three
base dimensions give $5 - 3 = 2$ dimensionless groups:
\[
\mathrm{Re} = \frac{\rho V c}{\mu}, \quad
\mathrm{Ma} = \frac{V}{a}.
\]
The structural result is
\[
C_{L} = C_{L}(\alpha, t/c, \mathrm{Re}, \mathrm{Ma}), \quad
C_{D} = C_{D}(\alpha, t/c, \mathrm{Re}, \mathrm{Ma}).
\]
Geometric similarity (same $\alpha$, same $t/c$, same planform)
is achieved by building a geometrically scaled model. Reynolds
matching and Mach matching are the two remaining requirements.

\paragraph{The full-scale design point.} A small UAV cruises at
$V_{p} = 40 \,\si{\meter\per\second}$ with chord
$c_{p} = 0.30 \,\si{\meter}$ at sea-level standard atmosphere
($\rho = 1.225 \,\si{\kilogram\per\meter\cubed}$, $\mu = 1.81
\times 10^{-5} \,\si{\pascal\second}$, $a = 340 \,\si{\meter\per
\second}$). The full-scale Reynolds and Mach are
$\mathrm{Re}_{p} = \rho V_{p} c_{p}/\mu = 1.225 \times 40 \times
0.30/(1.81 \times 10^{-5}) = 8.12 \times 10^{5}$ and
$\mathrm{Ma}_{p} = 40/340 = 0.118$.

\paragraph{A $1{:}20$ scale model in the same fluid.} The scale
model has chord $c_{m} = 0.015 \,\si{\meter}$. In ambient air, to
match Reynolds the test speed must be
\[
V_{m} = V_{p} \cdot \frac{c_{p}}{c_{m}}
  \cdot \frac{\rho_{p}}{\rho_{m}} \cdot \frac{\mu_{m}}{\mu_{p}}
  = 40 \times 20 \times 1 \times 1 = 800 \,\si{\meter\per\second}.
\]
The corresponding Mach is $\mathrm{Ma}_{m} = 800/340 = 2.35$,
fully supersonic. The model is in a different flow regime
(compressible, with shock waves) than the prototype (effectively
incompressible). The unconditional Reynolds match in ambient air
fails because Mach matching is impossible to satisfy at the same
geometric scale ratio.

\paragraph{Three working compromises.} Aerodynamicists resolve the
mismatch through one of three engineering choices.

The first is a \engterm{cryogenic tunnel} that lowers $\mu$ by
cooling the working fluid. NASA Langley's National Transonic
Facility (NTF) uses gaseous nitrogen at stagnation temperatures
down to $\sim 110 \,\si{\kelvin}$ and stagnation pressures up to
$\sim 9 \,\text{bar}$, raising $\rho$ and lowering $\mu$ together
to reach $\mathrm{Re}$ up to $1.4 \times 10^{8}$ on a chord of
$0.25 \,\si{\meter}$ at $\mathrm{Ma}$ matched to the prototype
\cite{web:ch02t3-nasa-ntf}. The NTF is the canonical answer to
the Reynolds-Mach matching problem for transport-aircraft testing.

The second is a \engterm{pressurised tunnel} that raises $\rho$
without changing $\mu$. The Variable Density Tunnel at NACA
Langley (1922--1941) ran at up to $20$ bar; raising $\rho$ by a
factor of $20$ at fixed $V$ and $c$ raises $\mathrm{Re}$ by the
same factor of $20$. The mismatch in Mach number remains, but
for incompressible-regime applications ($\mathrm{Ma} < 0.3$ in
the prototype) the Mach mismatch is acceptable.

The third is a \engterm{transition strip}: a roughened band glued
to the model's leading edge to trigger laminar-to-turbulent
transition at the same chordwise location it would occur on the
full-scale wing at the prototype's higher Reynolds. The strip
forces the boundary layer to behave as if $\mathrm{Re}$ were
higher, by removing the upstream laminar reach that the model's
low $\mathrm{Re}$ would otherwise preserve. Pope, Rae, and Barlow
discuss strip placement and grit sizing in detail
\cite{text:ch02t3-pope-experimental}.

\paragraph{A specialised facility example: icing.} For an iced-wing
flight test, the matching requirements include the liquid-water
content, the droplet median volume diameter, and the impingement
parameter (a dimensionless group of droplet inertia versus airflow
deflection). NASA Glenn's Icing Research Tunnel (IRT) supplies
the controlled liquid-water-content and droplet-size environment at
temperatures down to $-40 \,{}^{\circ}\mathrm{C}$, matching the
prototype's icing conditions rather than the prototype's Reynolds
and Mach
\cite{web:ch02t3-nasa-irt}. The lesson is that ``similarity'' has
as many dimensions as the problem has dimensionless groups; the
facility selects which dimensionless groups it can match and the
engineer accounts for the unmatched groups in interpretation.

\paragraph{The dimensional argument's working envelope.} For an
incompressible-regime UAV ($\mathrm{Ma}_{p} = 0.12$), a $1{:}20$
model in a pressurised tunnel at $20 \,\text{bar}$ matches
$\mathrm{Re}_{p}$ within a factor of two and ignores the
$\mathrm{Ma}$ mismatch with negligible error. For a transport
aircraft at cruise ($\mathrm{Ma}_{p} = 0.85$), the cryogenic
tunnel is the only working solution. For a hypersonic vehicle
($\mathrm{Ma}_{p} > 5$), the matching requirements expand to
include the Damk\"ohler number for finite-rate chemistry, and the
test facility set is even more specialised. Dimensional analysis
delivers the matching requirements; engineering judgement, the
budget, and the facility set deliver the achievable test.

\subsection{Why this matters}

Three reasons.

The first is \engterm{economy of experimentation}. To measure the
drag on a sphere across a wide range of conditions, one does not vary
$F$, $\rho$, $v$, $\mu$, and $D$ independently. One varies the
Reynolds number. A single curve $C_{D}(\mathrm{Re})$ summarises every
experiment for every fluid, every speed, every diameter. The
parameter space collapses from five dimensions to one. We will see
this collapse repeatedly: heat transfer reduces to a Nusselt-Reynolds
relationship, ship-towing scales by Froude number, vibration scales
by reduced frequency.

The second is \engterm{similarity in scale models}. Two geometrically
similar systems with the same governing dimensionless groups are
dynamically similar within the assumptions of the model. Wind-tunnel
data become useful when the relevant groups, geometry, boundary
conditions, and corrections are matched or accounted for. Aerodynamic
testing, naval architecture, hydraulic engineering, and
turbomachinery all rest on this discipline.

The third is \engterm{a check on the variable list}. If the
dimensionless groups do not correspond to anything physically
familiar, the variable list is probably wrong. If the experimental
data refuses to collapse onto a single curve $\Phi(\mathrm{Re})$, an
omitted variable is the most likely cause. The Buckingham pi theorem
turns the act of choosing variables into an auditable step.

\begin{mastery}
The choice of repeating variables is not unique. Any complete set of
$k$ variables that span the base dimensions independently will work,
and different choices give equivalent dimensionless groups. The
groups one writer obtains may look different from another's; they are
related by ratios and powers of the same dimensionless quantities. A
practised dimensional analyst chooses repeating variables that
produce groups with familiar physical interpretations (Reynolds,
Froude, Mach, Prandtl, Nusselt, Stokes, Strouhal, Bond, Grashof) so
that downstream readers do not have to translate.
\end{mastery}

\section{Failure: unit errors that cost money or lives}\pathtag{core}

The interface failure from Chapter 1 recurs in weaker forms: a test
instrument encodes the wrong geometry, a medical device hides the
meaning of a displayed dose, and a clinical order crosses from
notation to action with an unsafe abbreviation. We treat one
metrology-chain failure (Hubble), one display-interpretation case
(Therac-25), and the chronic class of medication unit errors, then
state the discipline.

\subsection{Hubble Space Telescope primary mirror, 1990}

The Hubble Space Telescope launched in April 1990 with a primary
mirror that was figured to the wrong shape. Images were blurred
because the mirror's outer edge was too flat by about $2 \,\si
{\micro\meter}$, an error large enough to ruin imaging across the
optical bandpass.

The error originated in the assembly of a reflective null corrector,
the optical-test instrument used during the mirror's polishing. A
field lens in the corrector was mispositioned by $1.3 \,\si{\milli
\meter}$ along the optical axis. The mispositioning resulted from a
misreading of a measurement target on a metering rod, in which the
end cap and the target feature were not properly distinguished. Two
auxiliary optical tests, a refractive null test and an inverse null
test, indicated a discrepancy at the time of fabrication; both were
set aside under cost and schedule pressure. The NASA Optical Systems
Board of Investigation found a flawed reflective-null-corrector
setup, missing independent verification, ignored auxiliary-test
evidence, and a decision chain that did not act on the early signals
\cite{acc:nasa-hubble-optical-systems-1990}.

\paragraph{Technical mechanism.} The reflective null corrector is an
optical-test instrument that places an aberrated wavefront on the
mirror surface to evaluate the figure during polishing. Its
internal element spacings determine the wavefront it imposes. A
$1.3 \,\si{\milli\meter}$ longitudinal misalignment of the field
lens produced a wavefront error that the polishing operation
faithfully matched, yielding a primary mirror that was too flat at
the outer edge by approximately $2 \,\si{\micro\meter}$ of optical
sag. The mirror was figured to high precision against the wrong
target.

\paragraph{Organisational context.} The mispositioning came from a
misreading of a metering rod whose end cap and an internal feature
were both candidate reference points. Two independent optical tests
flagged disagreement before launch: a refractive null test and an
inverse null test. Both were set aside under cost and schedule
pressure on the grounds that the reflective null corrector was the
primary instrument and the others were considered less reliable
checks. The NASA Optical Systems Board of Investigation documented
flawed setup, missing independent verification, ignored auxiliary
evidence, and a decision chain that did not act on early signals
\cite{acc:nasa-hubble-optical-systems-1990}.

\paragraph{Lesson for measurement.} A reference instrument is itself
a measurement chain. A null corrector's metering rod is a tool whose
indication is an interpretation of a measurement target; the
interpretation can be wrong; secondary checks exist to catch the
interpretation; the secondary checks must be honoured.
\Cref{fig:vol01:ch02:hubble-mirror} sketches the chain. The
1993 Servicing Mission 1 installed COSTAR to correct the wavefront
optically; subsequent Hubble instruments incorporated the correction
directly.

\input{volumes/01-quantity/ch02-units-and-dimensions/figures/fig-hubble-mirror.tex}

\subsection{Therac-25: unreadable displayed quantities, 1985--1987}

Therac-25 belongs first to software safety and medical-device
regulation. For this chapter, the relevant interface failure is
narrower: operators saw messages such as ``MALFUNCTION 54'' and dose
displays that did not tell them whether a dangerous overdose had
occurred. Leveson and Turner report six known overdose accidents
between June 1985 and January 1987, with deaths and serious
injuries, and describe the Tyler dose display as indicating an
underdose while the patient had received a massive overdose
\cite{paper:leveson-turner1993}.

Therac-25 belongs here only where the operator display reported a
quantity whose safety meaning the operator could not interpret. The
broader software, hazard-analysis, and regulatory mechanisms of the
case are taken up in Volume X.

\subsection{The chronic case: medication unit errors}

Medication-unit errors are a chronic class, not a single accident.
Patterns documented across pharmacy, nursing, emergency-medicine, and
medication-safety literature include microgram-milligram confusion,
millilitre-milligram confusion where concentration matters, unsafe
decimal-point notation, and insulin-unit abbreviation errors. The
mechanism is the same unit-confusion-at-an-interface pattern that the
named aerospace accidents illustrate, encountered many times per day
across many millions of patient interactions; specific magnitudes
should be cited from Joint Commission, ISMP, FDA, or peer-reviewed
medication-safety sources rather than asserted as house knowledge.

We return to medication errors in Volume X Chapter 9 under human
factors and operator error, and in Volume VI where medical devices
enter living systems.

\subsection{The discipline}

The discipline that prevents these failures is the same one we stated
in chapter 1, restated here in dimensional language.

\begin{principle}[Dimensional barrier at every interface]
Every quantity that crosses an interface, whether between humans,
between software systems, between hardware components, or between
documents, carries its unit explicitly. Conversions occur at named,
audited boundaries with both endpoints' units recorded. Dimensions
are checked at the interface, not assumed.
\end{principle}

The principle now has a name. At any interface, the receiving side
must know the quantity, unit, reference condition, range, and
conversion rule before it can safely act on the number.

\section{Project}\pathtag{core}

\begin{project}[Five derivations from dimensional analysis]
The reader applies the Buckingham pi machinery of section 2.5 to
five physical relationships and produces a small reference document
of derivations.

\textbf{Track}. Analysis. No physical work; no instruments. Pencil,
paper, and a calculator suffice; an optional plotting environment is
useful for one of the derivations but not required.
\textbf{Hazard class}: none.
\textbf{Tools}. Paper, pencil, calculator, optional plotting
software for one derivation.

\textbf{Procedure}. For each of the following five problems, the
reader (i) identifies the variables that the problem depends on,
(ii) writes each variable's dimensional signature, (iii) determines
$n$, $k$, and $n - k$, (iv) chooses repeating variables, (v) forms
the dimensionless groups, (vi) states the resulting structural
relationship, (vii) compares the structure to the published
functional form (where one exists), and (viii) names what
dimensional analysis cannot say about the prefactor or the
unmeasured function.

The five problems:

\begin{enumerate}[leftmargin=*]
  \item The period of a simple pendulum: variables $T, L, g, m$
    (mass should drop out).
  \item The terminal velocity of a sphere falling through a fluid:
    variables $v, \rho_{\text{sphere}} - \rho_{\text{fluid}}, \rho_
    {\text{fluid}}, g, D, \mu$. Identify the Stokes (low-Reynolds)
    and inertial-drag (high-Reynolds) limits separately.
  \item The time scale for a temperature pulse to diffuse through a
    slab: variables $\tau, L, \alpha$ (with $\alpha$ the thermal
    diffusivity).
  \item The free-fall time from rest through a height: variables
    $t, h, g$.
  \item The capillary rise of a wetting liquid in a thin tube:
    variables $h, \sigma, \rho, g, r, \theta$ (with $\sigma$ the
    surface tension, $r$ the tube radius, $\theta$ the contact
    angle).
\end{enumerate}

\textbf{Record}. Each derivation is one or two pages: variable list,
dimensional signatures, count, repeating variables, group formation,
conclusion, and comparison to the published formula. Assemble the
five derivations as a short dated reference document.

\textbf{Reflection (1500 words)}. The reader writes a 1500-word
narrative addressing:

\begin{itemize}[leftmargin=*]
  \item Which problem was hardest to choose variables for, and why.
  \item Which problem produced the most surprising dimensionless
    group.
  \item Which problem revealed an omitted variable when the result
    was compared to the published formula.
  \item Whether the choice of repeating variables affected the form
    of the group, and if so, how.
  \item One physical problem in the reader's own life or work where
    a dimensional argument would have caught a recent confusion.
\end{itemize}

\textbf{Deliverable}. The five derivations and the reflection. The
general editor's reference set of derivations is published online;
the reader is encouraged to compare \emph{after} completing their
own.

\textbf{Time}. Two to three hours per derivation, plus two to three
hours of writing. Total approximately fifteen hours, distributed
across two weeks.
\end{project}

\section{Exercises}

The exercises move from conversion and dimensional checks to
estimation, simulation, design, diagnosis, failure analysis, and
judgment. Calculation and derivation problems have full solutions;
open-ended problems have rubrics or reference derivations. Exercises
are tagged with their reader-path tier.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Convert $1\,\text{atm} = 101{,}325 \,\si{\pascal}$ to (a) bar,
(b) torr, (c) psi, (d) inches of mercury at $0\,\si{\degreeCelsius}$.
Show each conversion as multiplication by a ratio equal to one.
Comment on which form a meteorologist, a vacuum-system engineer, and
a US plumbing engineer would each use.

\paragraph{Solution.}
We perform each conversion as a multiplication by a ratio equal to
one, comment on the exactness of each ratio, and assign each form to
its working population.

\textit{Step 1, convert to bar.} The bar is defined as exactly
$10^{5} \,\si{\pascal}$. The conversion is
$1 \,\text{atm} = 101{,}325 \,\si{\pascal} \times (1 \,\text{bar} /
10^{5} \,\si{\pascal}) = 1.01325 \,\text{bar}$. The ratio is exact
by the definition of the bar.

\textit{Step 2, convert to torr.} The torr was defined as exactly
$1/760$ of a standard atmosphere when the standard atmosphere was
adopted in 1954. Therefore $1 \,\text{atm} = 760 \,\text{torr}$
exactly. In pascals, $760 \,\text{torr} \times 133.322 \,\si{\pascal}
/\text{torr} = 101{,}325 \,\si{\pascal}$ (the per-torr factor is
itself derived from $101{,}325 / 760$). The torr's relation to the
atmosphere is exact by definition; its relation to the pascal is a
ratio of exact quantities and is therefore also exact.

\textit{Step 3, convert to psi.} The pound-per-square-inch is defined
through the international pound and the international inch:
$1 \,\text{psi} = 1 \,\text{lbf}/(1 \,\text{in})^{2} = 4.44822
\,\si{\newton}/(0.0254 \,\si{\meter})^{2} = 6894.757 \,\si{\pascal}$
exactly (all input quantities are defined exactly). Then
$1 \,\text{atm} = 101{,}325 / 6894.757 = 14.696 \,\text{psi}$. The
ratio is exact in the sense that all defining quantities are exact;
the displayed decimal is rounded.

\textit{Step 4, convert to inches of mercury.} The inch of mercury
depends on the density of mercury, which depends on temperature.
The standard reference is mercury at $0 \,\si{\degreeCelsius}$ with
density $13{,}595.1 \,\si{\kilogram\per\meter\cubed}$ and standard
gravity $g_{0} = 9.80665 \,\si{\meter\per\second\squared}$. The
pressure of a one-inch column of mercury under these conditions is
$1 \,\text{inHg}_{0} = 0.0254 \,\si{\meter} \times 13{,}595.1
\,\si{\kilogram\per\meter\cubed} \times 9.80665 \,\si{\meter\per
\second\squared} = 3386.39 \,\si{\pascal}$. Then
$1 \,\text{atm} = 101{,}325 / 3386.39 = 29.92 \,\text{inHg}$. The
ratio is not exact: it depends on the chosen reference temperature
for mercury density, and meteorological practice may use
$15.6 \,\si{\degreeCelsius}$ (the US National Weather Service standard)
which gives a slightly different value.

\textit{Step 5, assign populations to working units.} Meteorologists
use millibars or hectopascals (numerically the same) in synoptic
forecasting because the millibar's relation to the atmosphere makes
the working pressures (typically $950$-$1050 \,\text{mbar}$) span a
range of meaningful magnitude. Vacuum engineers use torr or mbar at
low pressures (where the spans are orders of magnitude) and Pa for
high vacuum where the smallest pressures take values like
$10^{-10} \,\si{\pascal}$. US plumbing engineers use psi or inches of
mercury for water-system and HVAC work, following the convention of
the US building codes and the equipment-rating documentation.
Aerospace engineers report in psi or in hectopascals depending on
the audience and the regulatory regime.

\textit{Step 6, name the discipline.} The same physical pressure
carries four (or more) different numerical reports under the four
working conventions. The engineering response is to record the unit
explicitly on every entry of a calculation, and to perform any
conversion between systems by multiplying by a documented ratio
rather than by recalling a remembered factor. The most common error
is the silent conversion in a spreadsheet cell, where the unit-system
context is lost on copy-paste; the response is the unit-aware
spreadsheet column that carries the unit symbol alongside every
numerical value.
\end{exercise}

\begin{exercise}
A car accelerates from rest to $100 \,\si{\kilo\meter\per\hour}$ in
$8.0 \,\si{\second}$. Express the average acceleration in (a)
$\si{\meter\per\second\squared}$, (b) $g$ (multiples of standard
gravity), (c) $\text{ft}/\text{s}^{2}$. State which expression
makes the magnitude most legible.

\paragraph{Solution.}
We compute the acceleration in each unit, then evaluate the
legibility of each form against engineering intuition.

\textit{Step 1, normalise the speed.} The final speed of
$100 \,\si{\kilo\meter\per\hour}$ converts to SI by
$100 \times 10^{3} \,\si{\meter} / 3600 \,\si{\second} =
27.78 \,\si{\meter\per\second}$. The conversion factor
$1 \,\si{\kilo\meter\per\hour} = 1/3.6 \,\si{\meter\per\second}$ is
worth memorising; it appears in almost every vehicle-dynamics
calculation.

\textit{Step 2, compute (a) in SI.} For constant acceleration from
rest, $a = \Delta v / \Delta t = 27.78 / 8.0 = 3.47 \,\si{\meter\per
\second\squared}$. Three significant figures match the input
precision.

\textit{Step 3, compute (b) in multiples of $g$.} Standard gravity is
$g_{0} = 9.80665 \,\si{\meter\per\second\squared}$ exactly. The
acceleration is $3.47 / 9.80665 = 0.354 \,g$.

\textit{Step 4, compute (c) in imperial.} Using
$1 \,\text{ft} = 0.3048 \,\si{\meter}$ exactly,
$a = 3.47 \,\si{\meter\per\second\squared} \times (1 \,\text{ft}/
0.3048 \,\si{\meter}) = 11.4 \,\text{ft}/\text{s}^{2}$.

\textit{Step 5, evaluate legibility against body-felt calibration.}
Drivers and passengers have body-felt calibration for fractions of
$g$. A sharp braking event in a passenger car is approximately
$0.7$-$0.9 \,g$; a comfortable cornering load is $0.3$-$0.4 \,g$; a
gentle acceleration is $0.1$-$0.2 \,g$. The figure $0.354 \,g$ places
the vehicle on the boundary between gentle and brisk acceleration,
which matches a typical small car at full throttle. The SI form
$3.47 \,\si{\meter\per\second\squared}$ requires mental arithmetic
($\div 9.81$) to reach the same intuition; the imperial form
$11.4 \,\text{ft}/\text{s}^{2}$ requires the same arithmetic with a
less familiar denominator.

\textit{Step 6, name the engineering practice.} Automotive engineers
report acceleration in fractions of $g$ for human-factors purposes
(comfort, safety) and in $\si{\meter\per\second\squared}$ for
calculation purposes (drivetrain, braking, suspension). The two
representations are kept side-by-side in the engineering specification
so that human-factors review and physical-design review can read the
same number in the form each requires. The legibility of (b) is the
reason for the convention; the engineering trade-off is between
intuitive reading and dimensional homogeneity in the calculation.
\end{exercise}

\begin{exercise}
A reactor is rated for $300 \,\si{\kilo\watt}$ of thermal power.
Express this in (a) Btu per hour using the International Table Btu,
(b) mechanical horsepower, (c) tons of refrigeration. State each
conversion factor used and whether it is exact under that
convention.

\paragraph{Solution.}
We convert in each customary system, state the exactness of each
factor, and discuss where each conversion is used.

\textit{Step 1, convert to Btu/h.} The IT-Btu (International Table
Btu) was standardised in 1956 by the Fifth International Conference
on the Properties of Steam, fixing the relationship to SI as
$1 \,\text{Btu}_{\text{IT}} = 1055.05585 \,\si{\joule}$ exactly.
The hour-rate becomes $1 \,\text{Btu}_{\text{IT}}/\text{h} =
1055.05585 / 3600 = 0.29307107 \,\si{\watt}$ exactly. Then
$300 \,\si{\kilo\watt} = 3 \times 10^{5} \,\si{\watt} / 0.29307107
\,\si{\watt}/(\text{Btu}/\text{h}) \approx 1.024 \times 10^{6}
\,\text{Btu}/\text{h}$.

\textit{Step 2, convert to mechanical horsepower.} The mechanical
horsepower defined by James Watt is $1 \,\text{hp}_{\text{mech}} =
33{,}000 \,\text{ft}\cdot\text{lbf}/\text{min} =
550 \,\text{ft}\cdot\text{lbf}/\text{s}$. Converting through
$1 \,\text{ft}\cdot\text{lbf} = 1.35582 \,\si{\joule}$ (from
$0.3048 \,\si{\meter} \times 4.44822 \,\si{\newton}$, all exact),
$1 \,\text{hp}_{\text{mech}} = 550 \times 1.35582 = 745.69987
\,\si{\watt}$ exactly. Then $3 \times 10^{5} \,\si{\watt} / 745.7
\,\si{\watt}/\text{hp} \approx 402 \,\text{hp}$.

\textit{Step 3, convert to tons of refrigeration.} The
ton-of-refrigeration is the heat absorbed by melting one short ton
of ice ($2000 \,\text{lb}$) in $24 \,\si{\hour}$. The latent heat
of fusion of water is $144 \,\text{Btu}/\text{lb}$ (IT-Btu and
short-ton ice at $0 \,\si{\degreeCelsius}$). The total heat to melt
one ton is $2000 \times 144 = 2.88 \times 10^{5} \,\text{Btu}$; the
rate is $2.88 \times 10^{5} / 24 = 12{,}000 \,\text{Btu}/\text{h}$.
In watts, $12{,}000 \times 0.29307 = 3.517 \,\si{\kilo\watt}$. Then
$300 / 3.517 \approx 85.3 \,\text{tons of refrigeration}$.

\textit{Step 4, list each conversion's exactness.} (a) The Btu/h
conversion is exact under the IT-Btu definition; the IT-Btu is
itself a chosen standardisation of an older empirical definition,
so the factor is exact-by-convention rather than exact-by-physics.
(b) The mechanical-horsepower factor is exact in the foot-pound-
second system; the foot, the pound-force, and the
$745.69987 \,\si{\watt}$ value all carry exact-by-definition status.
(c) The ton-of-refrigeration is exact under the chosen ice-melting
definition; the underlying latent-heat-of-fusion figure is the
historical empirical value at $0 \,\si{\degreeCelsius}$ and is
standardised to ASHRAE practice rather than re-derived from
thermodynamic measurement.

\textit{Step 5, identify the domain of each unit.} The Btu/h is
the working unit in US heating, ventilating, and air-conditioning
practice; HVAC equipment is rated in Btu/h or in tons of
refrigeration (cooling) and in Btu/h or kW (heating). Mechanical
horsepower is the working unit in automotive, industrial-motor, and
small-engine practice in the US; the European convention uses
kilowatts. Tons of refrigeration is the working unit in commercial
and industrial cooling: a building-cooling-load specification of
``200 tons'' translates directly to chiller selection. None of the
three conversions appears in nuclear, aerospace, or process-
chemistry work, which standardised on SI decades ago.

\textit{Step 6, name the engineering practice.} A reactor's
$300 \,\si{\kilo\watt}$ thermal rating, the $1.024 \times 10^{6}
\,\text{Btu}/\text{h}$ HVAC-equivalent, and the $402 \,\text{hp}$
shaft equivalent are the same physical quantity expressed in three
populations' working units. The procurement document for a heat
exchanger that interfaces with this reactor carries the rating in
the unit system the supplier uses, with the SI equivalent in
parentheses; the engineering drawing carries the SI value with
the customary equivalent in parentheses. The principle is that
every interface carries both units explicitly, and the conversion
is verified at the interface by a second engineer rather than left
to mental arithmetic at the joint.
\end{exercise}

\begin{exercise}
A torque specification is given as $90 \,\si{\newton\meter}$ at a
fastener, $\pm 5\,\%$. Convert to foot-pounds-force and report the
tolerance band in both unit systems. Comment on which precision is
realisable with a standard click-type torque wrench.

\paragraph{Solution.}
We convert the central value and the tolerance band, evaluate the
wrench's contribution against the specification, and identify the
production practice.

\textit{Step 1, convert the central value.} Using the exact
conversion $1 \,\text{ft}\cdot\text{lbf} = 1.35582 \,\si{\newton
\meter}$, $90 \,\si{\newton\meter} \times (1 \,\text{ft}\cdot\text{
lbf}/1.35582 \,\si{\newton\meter}) = 66.38 \,\text{ft}\cdot\text{lbf}$,
which we round to $66.4 \,\text{ft}\cdot\text{lbf}$.

\textit{Step 2, convert the tolerance band.} The $\pm 5\,\%$
specification gives an SI tolerance band of $85.5$ to $94.5
\,\si{\newton\meter}$. Converting,
$85.5 \,\si{\newton\meter} \times 0.7376 = 63.07 \,\text{ft}\cdot
\text{lbf}$ and $94.5 \,\si{\newton\meter} \times 0.7376 = 69.71
\,\text{ft}\cdot\text{lbf}$. The imperial tolerance band is
$63.1$ to $69.7 \,\text{ft}\cdot\text{lbf}$ (the conversion
preserves the $\pm 5\,\%$ relative tolerance; the absolute band
width is $6.6 \,\text{ft}\cdot\text{lbf}$).

\textit{Step 3, characterise the standard click-type wrench.} A
production-grade click-type torque wrench is typically accurate to
$\pm 4\,\%$ of indicated reading from $20\,\%$ to $100\,\%$ of full
scale, per ASME B107.300. Below $20\,\%$ of full scale, the
accuracy degrades; the wrench should be selected so that the target
torque falls in the upper $40$-$80\,\%$ of its range. For a
$90 \,\si{\newton\meter}$ target, a wrench rated to
$\sim 135 \,\si{\newton\meter}$ full scale is appropriate.

\textit{Step 4, evaluate the instrument's contribution.} The
wrench's $\pm 4\,\%$ accuracy contributes $\pm 3.6 \,\si{\newton
\meter}$ to the achieved torque at the $90 \,\si{\newton\meter}$
target. The fastener's $\pm 5\,\%$ specification allows $\pm 4.5
\,\si{\newton\meter}$. The wrench accuracy alone consumes
approximately $80\,\%$ of the budget. The remaining $20\,\%$ must
absorb operator error: overshoot on the click (operators typically
push $1$-$2\,\%$ past the click), wrench-handle alignment (axial
torque is corrupted by side load), temperature drift (spring rate
shifts $\sim 0.1\,\%$ per kelvin), and joint friction (between bolt
head and joint surface, varies with thread lubrication and surface
finish). The $\pm 5\,\%$ specification is just within reach but
provides no margin against any of these contributors.

\textit{Step 5, name the corrective practice.} For tighter
tolerances ($\pm 2\,\%$ or better), the engineering practice
shifts to beam-type wrenches (which directly read the applied
torque on a deflection scale, eliminating the click overshoot) or
to transducer wrenches with digital readout (which apply the
calibration correction in firmware and remove the calibration
drift). For tightest tolerances ($\pm 1\,\%$ or better), preload
is measured directly via ultrasonic bolt-elongation or
strain-gauged bolts; the torque specification is replaced by a
preload specification at the joint.

\textit{Step 6, name the audit step.} The wrench's calibration
certificate is the documentation that survives a failure
investigation. The procedure document records the wrench's
identification number, the calibration date, the calibration
laboratory, the calibration uncertainty, and the wrench's recall
interval. The torque-applied value is recorded in the work-order
log at the time of application; the recording is countersigned by
a second technician. The audit trail is the engineering layer that
turns the $\pm 5\,\%$ specification into a verifiable claim rather
than a hope.
\end{exercise}

\begin{exercise}
A sound-pressure level is reported as $85 \,\si{\deci\bel}$ relative
to $20 \,\si{\micro\pascal}$. State (a) the actual sound pressure in
pascals, (b) the sound pressure of a level $20 \,\si{\deci\bel}$
higher, (c) the ratio between the two pressures.

\paragraph{Solution.}
We invert the decibel definition, compute each part, and connect the
logarithmic scale to the engineering convention.

\textit{Step 1, invert the decibel definition for pressure.} Sound-
pressure level is defined as
$L = 20 \log_{10}(p/p_{0})$ where $p_{0} = 20 \,\si{\micro\pascal}$
is the auditory threshold reference. Solving for $p$,
$p = p_{0} \cdot 10^{L/20}$.

\textit{Step 2, compute (a).} For $L = 85 \,\si{\deci\bel}$,
$p/p_{0} = 10^{85/20} = 10^{4.25} \approx 17{,}783$. Then
$p = 17{,}783 \times 20 \times 10^{-6} \,\si{\pascal} \approx 0.356
\,\si{\pascal}$. For comparison, atmospheric pressure is
$\sim 10^{5} \,\si{\pascal}$; the sound at $85 \,\si{\deci\bel}$
oscillates the air at $\sim 3.6 \times 10^{-6}$ of ambient pressure.

\textit{Step 3, compute (b).} A $20 \,\si{\deci\bel}$ increase
corresponds to a factor of $10^{20/20} = 10$ in pressure:
$p' = 10 \times 0.356 = 3.56 \,\si{\pascal}$. The pressure-level
increase of $105 \,\si{\deci\bel}$ is at the threshold of pain for
the human ear and corresponds to industrial machinery noise.

\textit{Step 4, compute (c) and identify the convention.} The
pressure ratio is exactly $10$ (a $20\,\si{\deci\bel}$ change on a
root-power quantity). The corresponding power ratio is
$10^{20/10} = 100$, because acoustic power scales as pressure
squared. The decibel is therefore $20 \log$ for amplitude quantities
(pressure, voltage, current) and $10 \log$ for power quantities
(acoustic intensity, electrical power). This factor-of-two
convention is the most-confused detail of decibel work; the
engineering practice is to label the decibel value with the
reference quantity explicitly: $\si{\deci\bel}$-SPL for sound
pressure, $\si{\deci\bel}$-A for A-weighted sound pressure,
$\si{\deci\bel}$m for milliwatts, $\si{\deci\bel}$V for volts.

\textit{Step 5, connect to human perception.} The logarithmic scale
was adopted because the human auditory system's response is roughly
logarithmic across at least six decades of pressure (from $20
\,\si{\micro\pascal}$ at the auditory threshold to $\sim 20
\,\si{\pascal}$ at the pain threshold). A linear pressure scale
would require six orders of magnitude on the working scale, with
the audible range mostly compressed near zero; the logarithmic
scale gives equal subjective steps approximately equal numerical
steps.

\textit{Step 6, name the arithmetic gotcha.} Decibel values do not
add arithmetically in the underlying pressure. Two
$85 \,\si{\deci\bel}$ sources combine to $85 + 3 = 88 \,\si{\deci\bel}$,
not to $170 \,\si{\deci\bel}$. The $+3 \,\si{\deci\bel}$ corresponds
to a doubling of power, which is the correct result for two
uncorrelated sources of equal intensity. The engineer who adds
decibel values arithmetically (treating them as power rather than
intensity additions) makes a factor-of-ten or larger error in the
result; the practice is to convert back to linear pressure or
intensity before summing, then convert the sum back to decibels.
\end{exercise}

\begin{exercise}
A blood-pressure reading is $120/80\,\text{mmHg}$. Convert
each value to (a) pascals, (b) kilopascals, (c) atmospheres. State
whether the reading is gauge or absolute, and identify which a
pressurised research application would require.

\paragraph{Solution.}
We convert each value, identify the gauge-vs-absolute distinction,
and state where each convention applies.

\textit{Step 1, identify the conversion factor.} One millimetre of
mercury at the standard reference (mercury at $0 \,\si{\degreeCelsius}$,
standard gravity) is $1 \,\text{mmHg} = 133.322 \,\si{\pascal}$
exactly under the chosen reference. The conversion is exact under
this convention; clinical practice does not require the temperature
correction because the cuff and the patient are at approximately
the same temperature, and the pressure measurement is differential
rather than absolute.

\textit{Step 2, convert the systolic value.}
$120 \,\text{mmHg} \times 133.322 \,\si{\pascal}/\text{mmHg} =
15{,}999 \,\si{\pascal} \approx 16.0 \,\si{\kilo\pascal}$. In
atmospheres, $16.0 \times 10^{3} / 101{,}325 = 0.158 \,\text{atm}$.

\textit{Step 3, convert the diastolic value.}
$80 \,\text{mmHg} \times 133.322 = 10{,}666 \,\si{\pascal} \approx
10.7 \,\si{\kilo\pascal} \approx 0.105 \,\text{atm}$.

\textit{Step 4, identify gauge vs absolute.} Clinical blood-pressure
readings are gauge pressures. The auscultatory method inflates a
cuff above arterial pressure, then deflates it slowly; the systolic
and diastolic values are the cuff pressures at which the Korotkoff
sounds first appear and disappear. The reference for the
measurement is atmospheric pressure on the outside of the cuff. The
displayed value is the pressure differential between arterial
pressure and atmosphere, not the absolute pressure inside the
artery.

\textit{Step 5, identify when absolute pressure matters.}
Pressurised research applications require absolute pressure
($p_{\text{abs}} = p_{\text{gauge}} + p_{\text{atm}}$) because the
physiological effects depend on the total pressure on vessel walls.
Examples include hyperbaric-chamber studies (where ambient pressure
is elevated $2$-$3 \,\text{atm}$), high-altitude physiology
preparations (where ambient pressure is reduced), and any
flow-through preparation in which the ambient pressure changes
during the experiment. For these cases the gauge reading must be
combined with a measurement of the ambient pressure to recover the
absolute value.

\textit{Step 6, name the working unit by domain.} Clinical practice
in most countries uses mmHg for blood pressure; the convention
predates SI adoption in medicine and has been retained for
continuity with the published clinical literature. Research
physiology increasingly reports in kPa, with mmHg in parentheses
for clinical readability. Engineering pressure work uses Pa or kPa
exclusively. The conversion to the working unit is performed once,
at the interface between domains, and is documented in the
experimental protocol.
\end{exercise}

\begin{exercise}
The standard molar enthalpy of combustion of glucose is approximately
$\Delta H \approx -2803 \,\si{\kilo\joule\per\mole}$ for products
$\mathrm{CO_2(g)}$ and $\mathrm{H_2O(l)}$ at standard conditions;
cite the thermochemical source. Convert this value to (a) calories
per mole and (b) electron-volts per molecule. State which form a
biochemist, a combustion engineer, and a particle physicist would
prefer.

\paragraph{Solution.}
We convert to each unit, identify the convention each domain uses,
and check the consistency of the three results against a fuel-energy
sanity check.

\textit{Step 1, convert to calories per mole.} The thermochemical
calorie is defined as exactly $1 \,\text{cal}_{\text{th}} = 4.184
\,\si{\joule}$. Then
$-2803 \,\si{\kilo\joule\per\mole} \times 10^{3} \,\si{\joule}/\si{
\kilo\joule} / 4.184 \,\si{\joule}/\text{cal} = -670{,}000 \,\text{cal}/
\text{mol} = -670 \,\text{kcal}/\text{mol}$.
The IT-calorie ($4.1868 \,\si{\joule}$) gives a slightly different
value; the thermochemical calorie is the convention in
biochemistry.

\textit{Step 2, convert to joules per molecule.} Dividing the molar
quantity by Avogadro's number $N_{A} = 6.02214 \times 10^{23}
\,\text{mol}^{-1}$,
$-2803 \times 10^{3} \,\si{\joule\per\mole} / 6.02214 \times 10^{23}
\,\text{mol}^{-1} = -4.655 \times 10^{-18} \,\si{\joule}/\text{molecule}$.

\textit{Step 3, convert to electron-volts per molecule.} The electron
volt is $1 \,\text{eV} = e \times 1 \,\text{volt} = 1.602176 \times
10^{-19} \,\si{\joule}$. Then
$-4.655 \times 10^{-18} / 1.602 \times 10^{-19} \approx -29.1
\,\text{eV}/\text{molecule}$.

\textit{Step 4, identify each domain's convention.} Biochemists use
kcal/mol because the historical biochemistry literature was tabulated
in this unit and the values of order $1$-$100$ are numerically
tractable. The kilocalorie is the food-calorie of nutrition labels:
the same calorie used in biochemistry. Combustion engineers use
kJ/mol or MJ/kg, with MJ/kg most informative for fuel comparisons:
glucose is approximately $\Delta H / M = 2803 \,\si{\kilo\joule\per
\mole} / 180.16 \,\si{\gram\per\mole} = 15.6 \,\si{\mega\joule\per
\kilogram}$, comparable to wood at $\sim 18 \,\si{\mega\joule\per
\kilogram}$ and below diesel at $\sim 45 \,\si{\mega\joule\per
\kilogram}$. Particle physicists use eV because their domain is
single-particle interactions where eV-scale is the natural unit: an
oxidation-reduction reaction at $30 \,\text{eV}$ per molecule
involves energies several orders of magnitude larger than an
ionisation or a chemical bond at $\sim 5$-$10 \,\text{eV}$.

\textit{Step 5, verify the energy density.} The combustion of glucose
delivers $15.6 \,\text{MJ/kg}$. Cellular respiration of glucose at the
ATP-yielding efficiency of $\sim 40\,\%$ delivers approximately
$6.2 \,\text{MJ/kg}$ as usable ATP energy. A $2000 \,\text{kcal}/
\text{day}$ human diet is approximately $8.4 \,\text{MJ/day}$, or
about $1.3 \,\text{kg}$ of glucose-equivalent food. The numbers are
mutually consistent.

\textit{Step 6, name the convention discipline.} The same energy is
expressed in three forms with different numerical prefactors. The
engineering response is to label every column of a calculation with
its unit, to record the convention (thermochemical vs IT-calorie, eV
vs joule, mass vs mole basis) on the first line of the table, and to
verify the dimensional consistency before propagating downstream.
Mistakes in chemistry-engineering interfaces often arise from
silent conversion between the mass-basis and mole-basis
representations; the discipline is to make the basis explicit at
every use.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Starting from $F = ma$, $\dd W = F \dd x$, and $K = \int F \dd x$,
derive the kinetic-energy expression $K = \tfrac{1}{2} m v^{2}$ for
a particle accelerated from rest. Verify that the dimensions of each
step are consistent.

\paragraph{Solution.}
We derive $K = \tfrac{1}{2} m v^{2}$ from the work-energy integral and
verify dimensional homogeneity at every step.

\textit{Step 1, write the differential work.} For a particle of mass
$m$ accelerated by a net force $F$ along $x$, the differential work
is $\dd W = F \dd x$.

\textit{Step 2, substitute Newton's second law.} With $F = m a$ and
$a = \dd v / \dd t$,
\[
\dd W = m a \dd x = m \frac{\dd v}{\dd t} \dd x.
\]

\textit{Step 3, change the integration variable.} Rearrange
$(\dd x / \dd t)$ as the speed $v$ so that
$\dd W = m \frac{\dd x}{\dd t} \dd v = m v \dd v$.
The substitution is valid because the particle's motion is
one-dimensional and $v = \dd x/\dd t$ along the path.

\textit{Step 4, integrate from rest to $v$.} The kinetic energy is the
total work done from rest:
\[
K = \int_{0}^{v} m v' \dd v' = \frac{1}{2} m v^{2}.
\]

\textit{Step 5, verify dimensional homogeneity of every step.} We
trace the dimensions through each substitution. (i) Force times
distance: $[F\dd x] = [M][L][T]^{-2} \times [L] = [M][L]^{2}[T]^{-2}$,
the energy dimension. (ii) Force as $m a$: $[m a \dd x] = [M] \times
[L][T]^{-2} \times [L] = [M][L]^{2}[T]^{-2}$, same. (iii) The
substitution $a \dd x = v \dd v$: $[v \dd v] = [L][T]^{-1} \times
[L][T]^{-1} = [L]^{2}[T]^{-2}$, so $[m v \dd v] = [M][L]^{2}[T]^{-2}$,
same. (iv) The integrated form: $[\tfrac{1}{2} m v^{2}] = [M] \times
([L][T]^{-1})^{2} = [M][L]^{2}[T]^{-2}$, same.

\textit{Step 6, name the engineering use.} The dimensional consistency
of every step is the audit trail that lets a working engineer trust
the derivation. A dimensional inconsistency at any intermediate step
signals a substitution error; the dimensions catch the error before
the algebra propagates it. The same energy dimension
$[M][L]^{2}[T]^{-2}$ appears throughout, and any term with a different
dimension that turns up in a subsequent calculation can be flagged
immediately as a sign that an algebraic mistake has been made.
\end{exercise}

\begin{exercise}
Show that, for a wave on a stretched string, the wave speed must
take the form $v = C\sqrt{T/\mu}$ where $T$ is the tension, $\mu$ is
the mass per unit length, and $C$ is a dimensionless constant. State
$C$ and the assumption that fixes it.

\paragraph{Solution.}
We list the dimensional signatures, solve the exponent system, derive
the prefactor from the wave equation, and identify the regime where
dimensional analysis stops.

\textit{Step 1, list the dimensional signatures.} Tension is a force,
so $[T] = [M][L][T]^{-2}$. Mass per unit length is
$[\mu] = [M][L]^{-1}$. Wave speed is $[v] = [L][T]^{-1}$. We seek
$v \propto T^{a} \mu^{b}$ where the constants $a$ and $b$ are fixed
by dimensional homogeneity.

\textit{Step 2, write the dimensional equation.} Matching dimensions
on the two sides,
\[
[L][T]^{-1}
= ([M][L][T]^{-2})^{a} ([M][L]^{-1})^{b}
= [M]^{a+b}\,[L]^{a-b}\,[T]^{-2a}.
\]

\textit{Step 3, solve the exponent system.} Equating exponents in each
base dimension,
\[
[M]: \quad a + b = 0, \qquad
[L]: \quad a - b = 1, \qquad
[T]: \quad -2a = -1.
\]
The $[T]$ equation gives $a = 1/2$; the $[M]$ equation then gives
$b = -1/2$; the $[L]$ equation closes consistently. Therefore
$v = C \sqrt{T/\mu}$ for a dimensionless constant $C$.

\textit{Step 4, derive the prefactor from the wave equation.} For a
uniform string of mass-per-length $\mu$ under tension $T$, the small-
amplitude transverse displacement $y(x, t)$ obeys the wave equation
\[
\frac{\partial^{2} y}{\partial t^{2}} = \frac{T}{\mu} \frac{\partial^{2}
y}{\partial x^{2}},
\]
derived by applying Newton's second law to a string element and
linearising the restoring force in the small-amplitude limit. The
general solution is $y(x, t) = f(x - vt) + g(x + vt)$ with
$v = \sqrt{T/\mu}$ exactly. The prefactor is $C = 1$.

\textit{Step 5, identify the regime of validity.} The wave equation is
the linearised approximation: the transverse displacement is small
compared to the wavelength, and the restoring force is strictly
proportional to displacement. Large-amplitude waves on a string have
two nonlinearities: the geometric nonlinearity (the tension changes
with displacement because the string stretches), and the constitutive
nonlinearity (the string's Young's modulus enters above the linear
strain regime). Both nonlinearities introduce additional dimensional
variables that dimensional analysis cannot capture without including
them in the variable list.

\textit{Step 6, state the engineering lesson.} Dimensional analysis
gave the structural form $v = C \sqrt{T/\mu}$, identifying that the
wave speed depends on $T$ and $\mu$ only through this ratio. The
prefactor $C = 1$ came from the differential equation, not from
dimensions. The regime of validity (small amplitude) is the
assumption that licenses the prefactor; outside that regime, the form
is still informative as a leading-order approximation, but the
prefactor and the variable list are both subject to correction. The
dimensional argument is the structural backbone; the differential
equation provides the prefactor and the regime boundary.
\end{exercise}

\begin{exercise}
Use dimensional analysis to derive the form of the period of a
spring-mass system in terms of the mass $m$ and the spring constant
$k$. State the dimensionless prefactor.

\paragraph{Solution.}
We list the variables, derive the dimensionless form, recover the
prefactor from the dynamical equation, and verify against a known
limit.

\textit{Step 1, list the dimensional signatures.} The mass is
$[m] = [M]$. The spring constant has dimensions of force per length,
$[k] = [F]/[L] = ([M][L][T]^{-2})/[L] = [M][T]^{-2}$. The period is
$[T_{p}] = [T]$. The candidates are $m$ and $k$; gravity does not
enter because the spring-mass oscillation is a balance between
inertia and elastic restoring force, neither of which depends on
$g$.

\textit{Step 2, write the dimensional equation.} Posit
$T_{p} \propto m^{a} k^{b}$. The dimensional homogeneity equation is
\[
[T] = [M]^{a} \cdot ([M][T]^{-2})^{b} = [M]^{a+b} \cdot [T]^{-2b}.
\]

\textit{Step 3, solve the exponent system.} Equating exponents,
\[
[M]: \quad a + b = 0, \qquad [T]: \quad -2b = 1.
\]
Solving, $b = -1/2$ and $a = 1/2$. Therefore
$T_{p} = C \sqrt{m/k}$ for a dimensionless $C$.

\textit{Step 4, derive the prefactor.} The dynamical equation for a
mass $m$ attached to a spring of constant $k$ in the absence of
damping is $m \ddot{x} = -k x$, or $\ddot{x} + (k/m) x = 0$. The
general solution is $x(t) = A \cos(\omega t + \phi)$ with
$\omega = \sqrt{k/m}$. The period is the time for one complete
cycle: $T_{p} = 2\pi / \omega = 2\pi \sqrt{m/k}$. The prefactor is
$C = 2\pi$.

\textit{Step 5, verify against a known limit.} For a $1 \,\si{\kilogram}$
mass on a spring of $k = 100 \,\si{\newton\per\meter}$,
$T_{p} = 2\pi \sqrt{1/100} = 2\pi \times 0.1 = 0.628 \,\si{\second}$.
A handheld pendulum or a small lab apparatus reproduces this period
within a few percent. The dimensional form's structural prediction
(period scales as $\sqrt{m}$ at fixed $k$; period scales as
$1/\sqrt{k}$ at fixed $m$) is verifiable in elementary lab
experiments and matches measurement to within the timing-resolution
floor.

\textit{Step 6, name the engineering use.} Dimensional analysis
recovers the structural form $T_{p} \propto \sqrt{m/k}$ from the
variable list alone; this is the same content the engineer would
use in scoping a vibration isolator's natural frequency before
committing to the full dynamical model. The prefactor of $2\pi$ is
known to anyone who has integrated the harmonic-oscillator
equation; the dimensional argument's value is in the scaling, not
in the prefactor. In practical isolator design, $f = 1/T_{p} =
(1/2\pi) \sqrt{k/m}$ is the working expression; the resulting
natural frequency is then placed sufficiently far from the
disturbing-frequency band to ensure isolation.
\end{exercise}

\begin{exercise}
Use dimensional analysis to derive the form of the orbital period of
a planet in terms of the orbital radius $r$, the central mass $M$,
and Newton's gravitational constant $G$. State the resulting form
and identify it as Kepler's third law.

\paragraph{Solution.}
We list dimensions, solve for the exponents, derive the prefactor
from the central-force problem, and verify against the Earth-Sun
system.

\textit{Step 1, list the dimensional signatures.} The orbital radius
is $[r] = [L]$. The central mass is $[M] = [M]$. Newton's
gravitational constant has dimensions
$[G] = [L]^{3}[M]^{-1}[T]^{-2}$. The signature of $G$ follows from
the law $F = GMm/r^{2}$: force divided by mass-squared per length-
squared, which simplifies to length-cubed per mass per time-squared.

\textit{Step 2, write the dimensional equation.} Posit
$T_{p} \propto r^{a} M^{b} G^{c}$. Matching dimensions,
\[
[T] = [L]^{a} \cdot [M]^{b} \cdot ([L]^{3}[M]^{-1}[T]^{-2})^{c}
= [L]^{a+3c} \cdot [M]^{b-c} \cdot [T]^{-2c}.
\]

\textit{Step 3, solve the exponent system.}
\[
[L]: \quad a + 3c = 0, \qquad
[M]: \quad b - c = 0, \qquad
[T]: \quad -2c = 1.
\]
The $[T]$ equation gives $c = -1/2$; the $[M]$ equation gives
$b = -1/2$; the $[L]$ equation gives $a = 3/2$. Therefore
$T_{p} = C \sqrt{r^{3} / (GM)}$, equivalently
$T_{p}^{2} = C^{2} r^{3} / (GM)$.

\textit{Step 4, identify Kepler's third law.} The relation
$T_{p}^{2} \propto r^{3}$ at fixed $GM$ is Kepler's third law in its
modern form. Kepler stated the empirical relation in 1619 for the
solar system; Newton derived it from his law of gravitation in 1687.
The dimensional argument recovers the structural form (square-cube
scaling) without solving the equations of motion.

\textit{Step 5, derive the prefactor.} For a circular orbit at
radius $r$ around a central mass $M$, the gravitational force
provides the centripetal acceleration:
$G M m / r^{2} = m v^{2}/r$. Solving for orbital speed,
$v = \sqrt{GM/r}$. The orbital period is the circumference divided
by the speed: $T_{p} = 2\pi r / v = 2\pi \sqrt{r^{3}/(GM)}$. The
prefactor is $C = 2\pi$ for a circular orbit. For an elliptical
orbit, the same relation holds with $r$ replaced by the semi-major
axis $a$; the result is $T_{p} = 2\pi \sqrt{a^{3}/(GM)}$, recovered
from solving the full central-force problem.

\textit{Step 6, verify against the Earth-Sun system.} With
$G = 6.674 \times 10^{-11} \,\si{\meter\cubed\per\kilogram\per\second
\squared}$, $M_{\odot} = 1.989 \times 10^{30} \,\si{\kilogram}$, and
$r = 1.496 \times 10^{11} \,\si{\meter}$ (one astronomical unit),
$T_{p} = 2\pi \sqrt{(1.496 \times 10^{11})^{3} / (6.674 \times 10^{-11}
\times 1.989 \times 10^{30})} \approx 3.156 \times 10^{7}\,\si{
\second} \approx 365.2 \,\text{days}$.
The result matches the Earth's sidereal year to four significant
figures. The dimensional structure and the prefactor together
reproduce the most familiar period in astronomy.
\end{exercise}

\begin{exercise}
Use dimensional analysis to derive the form of the speed of sound in
an ideal gas in terms of the bulk modulus $K$ (units of pressure)
and the density $\rho$. Compare to the published form $c = \sqrt
{\gamma p / \rho}$ for an ideal gas and identify the dimensional and
physical content of $\gamma$.

\paragraph{Solution.}
We derive the form by dimensional analysis, identify the bulk modulus
appropriate for sound, recover the ideal-gas expression, and verify
against air at room temperature.

\textit{Step 1, list the dimensional signatures.} Bulk modulus has
the dimensions of pressure: $[K] = [M][L]^{-1}[T]^{-2}$. Density is
$[\rho] = [M][L]^{-3}$. Sound speed is $[c] = [L][T]^{-1}$.

\textit{Step 2, write the dimensional equation.} Posit $c \propto
K^{a} \rho^{b}$. Matching,
\[
[L][T]^{-1}
= ([M][L]^{-1}[T]^{-2})^{a} \cdot ([M][L]^{-3})^{b}
= [M]^{a+b} \cdot [L]^{-a-3b} \cdot [T]^{-2a}.
\]

\textit{Step 3, solve the exponent system.}
\[
[M]: \quad a + b = 0, \qquad
[L]: \quad -a - 3b = 1, \qquad
[T]: \quad -2a = -1.
\]
The $[T]$ equation gives $a = 1/2$; the $[M]$ equation gives
$b = -1/2$; the $[L]$ equation $-1/2 + 3/2 = 1$ closes consistently.
Therefore $c = C \sqrt{K/\rho}$ for a dimensionless $C$.

\textit{Step 4, derive the prefactor from the wave equation.} The
linearised acoustic-wave equation for small-amplitude pressure
fluctuations in a continuous medium is
\[
\frac{\partial^{2} p'}{\partial t^{2}} = \frac{K}{\rho} \nabla^{2} p',
\]
where $p'$ is the pressure perturbation. The general solution
admits travelling waves at speed $c = \sqrt{K/\rho}$, so $C = 1$.

\textit{Step 5, identify the appropriate $K$ for an ideal gas.} Sound
propagation in a gas involves compression at the wave frequency, on
a time-scale much shorter than the thermal-equilibration time. The
compression is therefore adiabatic, not isothermal. For an
ideal gas under adiabatic compression, $p V^{\gamma} = \text{const}$,
where $\gamma = c_{p}/c_{v}$ is the ratio of specific heats. The
adiabatic bulk modulus is $K_{\text{ad}} = \gamma p$ (compare
$K_{\text{iso}} = p$ for isothermal compression). Substituting,
$c = \sqrt{\gamma p / \rho}$. For air at room conditions,
$\gamma = 1.40$.

\textit{Step 6, verify and identify the dimensional limit.} For air
at $20 \,\si{\degreeCelsius}$, $p = 101{,}325 \,\si{\pascal}$,
$\rho = 1.204 \,\si{\kilogram\per\meter\cubed}$. Computing,
$c = \sqrt{1.40 \times 101{,}325 / 1.204} = \sqrt{117{,}851}
\approx 343 \,\si{\meter\per\second}$. The tabulated value at the
same conditions is $343.2 \,\si{\meter\per\second}$. The agreement
confirms the derivation. The factor $\gamma$ cannot be derived by
dimensional analysis because it is intrinsically dimensionless;
its value arises from the molecular degrees of freedom (the kinetic
theory result $\gamma = (f + 2)/f$ for a gas with $f$ degrees of
freedom per molecule, giving $\gamma = 5/3$ for monatomic gases and
$\gamma = 7/5$ for diatomic gases). Dimensional analysis cannot
derive dimensionless prefactors; it can only identify them as the
points where physics outside the dimensional structure determines
the answer.
\end{exercise}

\begin{exercise}
For laminar flow in a horizontal pipe, the volumetric flow rate $Q$
is known to depend on the pressure gradient $\dd P / \dd x$, the
viscosity $\mu$, and the radius $R$. Use dimensional analysis to
derive the structural form of the Hagen--Poiseuille relationship and
identify the prefactor that dimensional analysis cannot give.

\paragraph{Solution.}
We list the dimensional signatures, solve for the exponents,
identify Hagen-Poiseuille, derive the prefactor from the parabolic
velocity profile, and verify against a worked case.

\textit{Step 1, list the dimensional signatures.} The pressure
gradient is force per area per length:
$[\dd P/\dd x] = [M][L]^{-1}[T]^{-2} / [L] = [M][L]^{-2}[T]^{-2}$.
Dynamic viscosity is $[\mu] = [M][L]^{-1}[T]^{-1}$. The pipe radius
is $[R] = [L]$. The volumetric flow rate is
$[Q] = [L]^{3}[T]^{-1}$.

\textit{Step 2, write the dimensional equation.} Posit
$Q \propto (\dd P/\dd x)^{a} \mu^{b} R^{c}$. Matching,
\[
[L]^{3}[T]^{-1}
= [M]^{a+b} \cdot [L]^{-2a - b + c} \cdot [T]^{-2a - b}.
\]

\textit{Step 3, solve the exponent system.}
\[
[M]: \quad a + b = 0, \qquad
[L]: \quad -2a - b + c = 3, \qquad
[T]: \quad -2a - b = -1.
\]
From the $[M]$ equation, $b = -a$. Substituting into the $[T]$
equation: $-2a - (-a) = -a = -1$, so $a = 1$. Then $b = -1$. The
$[L]$ equation gives $c = 3 + 2a + b = 3 + 2(1) + (-1) = 4$.
Therefore $Q = C (\dd P/\dd x) R^{4} / \mu$.

\textit{Step 4, identify the Hagen-Poiseuille relation.} The
structural form $Q \propto R^{4}$ is one of the strongest scaling
results in fluid mechanics: a doubling of pipe radius increases the
volumetric flow rate by a factor of sixteen at fixed pressure
gradient. This is why veins are wider than arterioles, why pipe
sizing matters more than length in head-loss calculations, and why
the human lung's bronchioles are graded in radius to balance flow
distribution across alveoli. The structural prediction is the
working result before any prefactor.

\textit{Step 5, derive the prefactor from the parabolic profile.}
For steady, fully-developed laminar flow in a horizontal circular
pipe, the Navier-Stokes equations with no-slip at the wall yield a
parabolic velocity profile: $v(r) = (1/4\mu) (-\dd P/\dd x)(R^{2} -
r^{2})$. Integrating over the pipe cross-section,
\[
Q = \int_{0}^{R} v(r) \cdot 2\pi r \dd r
= \frac{2\pi}{4\mu}(-\dd P/\dd x) \int_{0}^{R} (R^{2} - r^{2}) r \dd r.
\]
The integral evaluates to $R^{4}/4$, so
$Q = (\pi/8\mu)(-\dd P/\dd x) R^{4} =
-\pi R^{4}/(8\mu) \cdot \dd P/\dd x$. The prefactor is
$C = -\pi/8$ (the sign reflects that flow goes from high pressure
to low pressure; the magnitude is $\pi/8 \approx 0.393$).

\textit{Step 6, verify against a worked case.} Consider water at
$20 \,\si{\degreeCelsius}$ flowing through a horizontal pipe of
radius $5 \,\si{\milli\meter}$ and length $1 \,\si{\meter}$ under
a pressure drop of $1 \,\si{\kilo\pascal}$. The viscosity is
$\mu = 10^{-3} \,\si{\pascal\second}$. The pressure gradient is
$\dd P/\dd x = -1000 \,\si{\pascal\per\meter}$. Then
$Q = \pi (0.005)^{4} \times 1000 / (8 \times 10^{-3}) =
\pi \times 6.25 \times 10^{-10} \times 1000 / 8 \times 10^{-3} =
\pi \times 6.25 \times 10^{-10} \times 1.25 \times 10^{5}
= \pi \times 7.81 \times 10^{-5} \approx 2.45 \times 10^{-4}
\,\si{\meter\cubed\per\second} \approx 245 \,\si{\milli\liter\per
\second}$. The Reynolds number for this flow is
$\mathrm{Re} = \rho v D / \mu = 1000 \times v \times 0.01 / 10^{-3}$
where $v = Q / (\pi R^{2}) = 2.45 \times 10^{-4} / (\pi \times
2.5 \times 10^{-5}) = 3.12 \,\si{\meter\per\second}$, giving
$\mathrm{Re} \approx 31{,}200$. This exceeds the laminar limit of
$\mathrm{Re} \approx 2300$, so the assumed laminar profile fails.
Hagen-Poiseuille applies only when laminar flow is verified; the
dimensional analysis result is necessary but not sufficient. A
working calculation always checks the Reynolds number before
applying the formula.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the dimensionless Mach number for a passenger jet cruising
at $250 \,\si{\meter\per\second}$ at an altitude where the speed of
sound is approximately $295 \,\si{\meter\per\second}$. State the
range of regimes (subsonic, transonic, supersonic) and place the
flight in one of them.

\paragraph{Reference estimate.}
$\mathrm{Ma} = v/c = 250/295 \approx 0.85$. Regimes: subsonic
$\mathrm{Ma} < 0.8$; transonic $0.8 \leq \mathrm{Ma} \lesssim 1.2$;
supersonic $1.2 \lesssim \mathrm{Ma} < 5$; hypersonic $\mathrm{Ma}
> 5$. This passenger jet is at the high end of the transonic regime,
which is where modern commercial cruise sits: high enough for fuel
efficiency, low enough to avoid the substantial wave-drag penalty
that begins as flow over the wings reaches local supersonic speed.
\end{exercise}

\begin{exercise}
A swimming athlete moves at about $2 \,\si{\meter\per\second}$
through water of density $1000 \,\si{\kilogram\per\meter\cubed}$ and
viscosity $\mu = 10^{-3} \,\si{\pascal\second}$. The athlete's
characteristic length is about $2 \,\si{\meter}$. Estimate the
Reynolds number. State whether laminar viscous theory is appropriate
or whether turbulent drag dominates.

\paragraph{Reference estimate.}
$\mathrm{Re} = \rho v L / \mu = 10^{3} \times 2 \times 2 / 10^{-3}
= 4 \times 10^{6}$. Turbulent: above $\mathrm{Re} \sim 10^{5}$ on
streamlined bodies, the boundary layer is turbulent across most of
the wetted surface. The athlete's drag is overwhelmingly form drag
plus turbulent skin friction; Stokes-flow laminar theory
($\mathrm{Re} < 1$) does not apply by nine orders of magnitude. The
swimming community's biomechanical models address turbulent
boundary-layer separation, wave drag at the surface, and propulsive
efficiency, not laminar viscous flow.
\end{exercise}

\begin{exercise}
Estimate the Froude number $\mathrm{Fr} = v / \sqrt{g L}$ for a
freight ship at cruising speed and identify the regime
(displacement, semi-displacement, planing) it places the ship in.
State the source of your speed and length figures.

\paragraph{Reference estimate.}
A typical large container ship: $L \approx 350 \,\si{\meter}$,
$v \approx 12 \,\si{\meter\per\second}$ (about $24 \,\text{kn}$).
$\mathrm{Fr} = 12 / \sqrt{9.81 \times 350} \approx 12 / 58.6 \approx
0.205$. Regime: displacement (below $\mathrm{Fr} \approx 0.4$).
Container ships cannot achieve semi-displacement or planing without
prohibitive power; their hull form is optimised for displacement
operation. Source: Maersk Triple-E class and equivalent; published
cruise data widely available in marine-trade and IMO sources.
\end{exercise}

\begin{exercise}
Estimate the Biot number $\mathrm{Bi} = h L / k$ for a metal cooking
pot of wall thickness $L = 4 \,\si{\milli\meter}$, thermal
conductivity $k = 50 \,\si{\watt\per\meter\per\kelvin}$, with a
convective heat-transfer coefficient $h \approx 10 \,\si{\watt\per
\meter\squared\per\kelvin}$. Comment on whether a lumped-thermal
analysis is appropriate.

\paragraph{Reference estimate.}
$\mathrm{Bi} = h L / k = 10 \times 0.004 / 50 = 8 \times 10^{-4}$.
Far below the lumped-capacitance threshold of $\mathrm{Bi} < 0.1$;
the pot wall has nearly uniform temperature through its thickness
during normal cooking. The lumped-thermal model is appropriate: the
internal conduction is so fast compared to the surface convection
that spatial gradients within the wall are negligible. The model
fails for fast transient processes (high $h$, e.g., quenching) or
thick low-conductivity walls (e.g., a ceramic crock pot with thick
walls).
\end{exercise}

\begin{exercise}
The Prandtl number $\mathrm{Pr} = \nu / \alpha$ is the ratio of
momentum diffusivity to thermal diffusivity. Estimate it for (a)
air, (b) water, (c) liquid mercury, (d) engine oil. State which
fluid most resists thermal diffusion relative to momentum.

\paragraph{Reference estimate.}
Approximate values near room temperature:
(a) Air $\mathrm{Pr} \approx 0.71$. Momentum and heat diffuse at
comparable rates; this is the kinetic-theory result for ideal gases.
(b) Water $\mathrm{Pr} \approx 7$. Heat diffuses much more slowly
than momentum.
(c) Liquid mercury $\mathrm{Pr} \approx 0.025$. Very high thermal
conductivity (metallic electron transport) makes heat diffuse much
faster than momentum.
(d) Engine oil $\mathrm{Pr} \approx 1000$-$10{,}000$ depending on
temperature. Very high viscosity and modest thermal conductivity
make heat diffusion extremely slow relative to momentum.
The engine oil most resists thermal diffusion relative to momentum;
this is why oil-cooled systems require flow to transfer heat rather
than relying on conduction.
\end{exercise}

\begin{exercise}
Estimate the surface tension of water from the capillary-rise
formula $h = 2\sigma\cos\theta / (\rho g r)$ given that water rises
about $30 \,\si{\milli\meter}$ in a glass tube of radius $0.5 \,\si
{\milli\meter}$. Compare to a cited room-temperature value for water,
about $\sigma \approx 72 \,\si{\milli\newton\per\meter}$ near
$20 \,\si{\degreeCelsius}$ to $25 \,\si{\degreeCelsius}$. This is a
material property at a stated temperature, not a current-as-of
statistic.

\paragraph{Reference estimate.}
Solve for $\sigma$: $\sigma = h \rho g r / (2 \cos\theta)$. For
water on clean glass, $\cos\theta \approx 1$ (wetting; contact
angle near zero). Substituting: $\sigma = 0.030 \times 1000 \times
9.81 \times 0.0005 / 2 \approx 0.0736 \,\si{\newton\per\meter}
= 73.6 \,\si{\milli\newton\per\meter}$. Compares to the published
$\approx 72 \,\si{\milli\newton\per\meter}$ at room temperature to
within $2\,\%$; the agreement is consistent with the
contact-angle and capillary-rise measurement uncertainty. A glass
tube not freshly cleaned would have larger $\theta$ and the
estimate would be biased low.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Generate $10^{4}$ values of $\mathrm{Re}$ logarithmically spaced
between $10^{-2}$ and $10^{6}$. Plot a named published correlation
for the drag coefficient of a sphere, $C_{D}(\mathrm{Re})$, on a
log-log axis; cite the correlation and state its Reynolds-number
range before plotting. Identify (a) the Stokes-flow regime where
$C_{D} \approx 24 / \mathrm{Re}$, (b) the constant-$C_{D}$ regime
around $\mathrm{Re} \approx 10^{4}$, (c) the drag crisis around
$\mathrm{Re} \approx 3 \times 10^{5}$.

\paragraph{Solution sketch.} See \texttt{code/drag\_curve.py} for the
Morsi-Alexander piecewise correlation (\cite{paper:morsi-alexander-1972}),
valid $10^{-1} < \mathrm{Re} < 5 \times 10^{4}$. (a) At $\mathrm{Re}
= 0.1$, $C_{D} \approx 240$, matching Stokes. (b) At $\mathrm{Re}
= 10^{4}$, $C_{D} \approx 0.4$, the constant Newton-regime value.
(c) The drag crisis is not captured by Morsi-Alexander; one needs
the high-Re correlations (e.g., Achenbach 1972) to see the dip
near $\mathrm{Re} \approx 3 \times 10^{5}$ from laminar-to-turbulent
boundary-layer transition that delays separation.
\end{exercise}

\begin{exercise}
Numerically verify the Buckingham pi result for the simple-pendulum
period: simulate the pendulum's swing for at least ten different
combinations of $L$ and $g$, measure the period in each case, and
demonstrate that $T\sqrt{g/L}$ is constant within numerical error.
Plot $T$ versus $\sqrt{L/g}$ and identify the slope as $2\pi$.

\paragraph{Solution sketch.} See \texttt{code/pendulum\_pi.py}.
For 10 combinations of $L$ (0.1 to 10 m) and $g$ (3 to 30 m/s$^2$),
integrate the small-angle equation $\ddot\theta + (g/L)\theta = 0$
with $\theta_0 = 0.05$ rad and measure the period from zero-
crossings. Across all 10 runs, $T\sqrt{g/L}$ equals $2\pi$ to 4
significant figures. The slope of $T$ vs $\sqrt{L/g}$ is exactly
$2\pi \approx 6.2832$. Large-amplitude runs ($\theta_0 = 0.5$ rad)
show the elliptic-integral correction with $T$ growing by $\sim
1.5\,\%$, confirming that the dimensional result captures the
small-amplitude limit only.
\end{exercise}

\begin{exercise}
For the wave equation on a string, $v = \sqrt{T/\mu}$, write a
simulation that solves the one-dimensional wave equation for at
least three combinations of $T$ and $\mu$, measures the wave speed,
and confirms the dimensional-analysis prediction. Plot measured speed
versus $\sqrt{T/\mu}$.

\paragraph{Solution sketch.} See \texttt{code/wave\_speed.py}.
Finite-difference solution of $\partial_t^2 u = (T/\mu) \partial_x^2 u$
on a string of length 1 m with three $(T, \mu)$ combinations:
(10 N, 0.01 kg/m), (40 N, 0.04 kg/m), (90 N, 0.01 kg/m). Measured
wave speeds from tracking a Gaussian pulse: $\approx 31.6$, $\approx
31.6$, $\approx 94.9 \,\si{\meter\per\second}$ respectively, matching
the predicted $\sqrt{T/\mu}$ values within $<1\,\%$. The first two
runs have identical $T/\mu$ ratios despite different absolute
values, confirming that the dimensionless group governs propagation.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a wind-tunnel experiment to measure the drag of a $1{:}10$
scale model of a car. State the wind-tunnel speed required to
reproduce the highway-speed Reynolds number, the assumptions you
make about the working fluid, and the accuracy required of the speed
measurement.

\paragraph{Rubric.} Full-scale: $L \approx 4 \,\si{\meter}$, $v
\approx 30 \,\si{\meter\per\second}$ (highway), giving
$\mathrm{Re} = \rho v L / \mu \approx 1.2 \times 30 \times 4 /
(1.8 \times 10^{-5}) \approx 8 \times 10^{6}$. Model: $L' = 0.4 \,
\si{\meter}$, $10\times$ smaller. Same fluid (air). To preserve
Reynolds number, $v' = 10 v = 300 \,\si{\meter\per\second}$, which
exceeds Mach 0.9 and brings in compressibility (Mach-number mismatch).
This is the standard wind-tunnel scaling difficulty: full Reynolds
matching at small scale requires Mach-number-violating speeds.
Engineering compromises: accept partial Reynolds match (e.g., $v'
\approx 50 \,\si{\meter\per\second}$, $\mathrm{Re}' \approx 1.3
\times 10^{6}$, an order of magnitude below full-scale but in a
roughly self-similar regime); use pressurised tunnels (raising
$\rho$); apply post-test Reynolds-number corrections. Speed
measurement accuracy: $\pm 1\,\%$ produces drag-coefficient
uncertainty $\sim 2\,\%$ (since drag scales as $v^2$); standard
wind-tunnel pitot-static accuracy is comfortably within this.
\end{exercise}

\begin{exercise}
Design a measurement plan to determine the thermal conductivity of
an unknown material from a transient-heating experiment. State the
quantities to be measured, the dimensionless groups they must
collapse onto, and the expected precision of the conductivity result.

\paragraph{Rubric.} The transient-hot-plane or hot-disk method
applies a step heat flux $q$ at the centre of a sample and records
temperature $T(t)$ at a known offset. Quantities: $q$ (W),
specimen geometry $L$ (m), heating duration, $T(t)$ time series,
specimen density $\rho$ and specific heat $c_p$ (or measure
thermal diffusivity directly and infer $k = \rho c_p \alpha$). The
governing dimensionless group is the Fourier number $\mathrm{Fo}
= \alpha t / L^{2}$; temperature response data collapses onto a
universal curve when plotted against $\mathrm{Fo}$. Expected
precision: $\pm 2$-$5\,\%$ on $k$ with proper instrumentation;
dominant error sources are contact thermal resistance at the
sensor interface and finite-sample boundary effects. The TPS
(transient plane source) standard ISO 22007-2 codifies the
implementation.
\end{exercise}

\begin{exercise}
Design a flow-meter calibration scheme based on dimensional
similarity rather than absolute calibration. State the variables,
the dimensionless groups, and the calibration's range of validity.
Identify a real flow-meter type to which the scheme applies.

\paragraph{Rubric.} Orifice-plate flow meters obey a similarity law
in $\mathrm{Re}$ and $\beta$ (the diameter ratio). The discharge
coefficient $C_{d}(\mathrm{Re}, \beta)$ is universal for geometrically
similar installations. A laboratory calibration on water at
$\mathrm{Re} = 10^{5}$ predicts the meter's behaviour on a different
fluid at the same $\mathrm{Re}$ and same $\beta$, eliminating the
need for fluid-specific calibration. Range of validity: typically
$\mathrm{Re} > 5 \times 10^{3}$ where $C_{d}$ is nearly Reynolds-
independent; outside this, additional dimensionless groups (surface
roughness, beta) become significant. The ISO 5167 standard codifies
this scheme for orifice plates, nozzles, and Venturi tubes; the
same principle applies to other geometrically-similar restriction
meters.
\end{exercise}

\begin{exercise}
A new mobile phone battery is to be tested for thermal performance.
Specify the dimensionless groups that govern its temperature rise
during fast charging, the test conditions that span those groups,
and the criterion by which a test is judged equivalent to a different
test on a different cell geometry.

\paragraph{Rubric.} The relevant groups for a charging Li-ion cell:
the dimensionless C-rate (charging current normalised by cell
capacity, giving a time-scale ratio between charging and the
cell's thermal relaxation), the Biot number $\mathrm{Bi} = h L / k$
(internal-conduction vs surface-cooling balance), the Fourier number
$\mathrm{Fo} = \alpha t / L^{2}$ (transient-heat-conduction
similarity), and an aspect-ratio group $L/D$ (cell geometry).
Span the C-rate from $0.5\mathrm{C}$ to $3\mathrm{C}$; span
$\mathrm{Bi}$ by varying cooling-fluid velocity; span $\mathrm{Fo}$
by varying test duration. Two different cell geometries are
thermally equivalent (in the sense that their temperature rises
scale the same way) when their $(C, \mathrm{Bi}, \mathrm{Fo},
L/D)$ are matched. The cell-level safety standard IEC 62133 and
the OEM-specific test protocols apply this principle.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
The following equation appears in a draft engineering report:
\[
\sigma = \frac{F^{2} L}{A E},
\]
with $\sigma$ a stress, $F$ a force, $L$ a length, $A$ an area, and
$E$ a modulus. State whether the equation is dimensionally correct.
If not, identify the error and propose a corrected form.

\paragraph{Solution.}
The dimensional check on engineering equations is the chapter's
primary working diagnostic. The solution applies the check to the
draft equation, identifies the mismatch, infers the intended
expression, and proposes a corrected form.

\emph{Step 1: state the dimensions of the left-hand side.} Stress
$\sigma$ is force per unit area, with dimensions
\[
[\sigma] = [F]/[L]^{2} = [M][L]^{-1}[T]^{-2}.
\]

\emph{Step 2: assemble the dimensions of the right-hand side.}
$F^{2}L/(AE)$:
\begin{itemize}[leftmargin=*]
  \item $[F]^{2} = [M]^{2}[L]^{2}[T]^{-4}$
  \item $[L] = [L]$
  \item $[A] = [L]^{2}$
  \item $[E] = [F]/[L]^{2} = [M][L]^{-1}[T]^{-2}$ (modulus has the
    dimensions of stress, since it relates stress to strain and
    strain is dimensionless)
\end{itemize}
The combined dimensions are
\[
\frac{[F]^{2}[L]}{[A][E]} =
\frac{[M]^{2}[L]^{3}[T]^{-4}}{[L]^{2}[M][L]^{-1}[T]^{-2}} =
\frac{[M]^{2}[L]^{3}[T]^{-4}}{[M][L][T]^{-2}} = [M][L]^{2}[T]^{-2}.
\]

\emph{Step 3: compare to the required dimensions.} The right-hand
side has dimensions $[M][L]^{2}[T]^{-2}$, which are those of
energy. The required dimensions are $[M][L]^{-1}[T]^{-2}$ (stress).
The mismatch is by $[L]^{3}$, equivalently by a factor of volume.
The equation is dimensionally wrong.

\emph{Step 4: infer the intended expression.} The equation
resembles the elastic-elongation form $\delta = FL/(AE)$, which is
dimensionally correct ($[L][L]/([L]^{2}[F]/[L]^{2}) = [L]$). The
draft has $F^{2}$ where elongation has $F$, suggesting the report
wrote elongation and then "squared" by accident, perhaps in a
substitution or units-conversion step. Alternatively, the writer
may have confused stress with strain energy (stress times strain
times volume), which would give the dimensions of energy on the
right.

\emph{Step 5: propose a corrected form.} Three working corrections,
depending on the context the report is trying to express:
\begin{itemize}[leftmargin=*]
  \item Axial stress under load $F$ over cross-section $A$:
    $\sigma = F/A$.
  \item Axial strain from elongation $\delta$ over length $L$:
    $\varepsilon = \delta/L$.
  \item Stress-strain relation (Hooke's law): $\sigma = E\varepsilon$.
  \item Elongation under load: $\delta = FL/(AE)$.
\end{itemize}
The right-hand side of the draft equation, $F^{2}L/(AE)$, has the
dimensions of strain energy: $U = F^{2}L/(2AE)$ is the elastic
strain energy stored in a uniaxially loaded bar.

\emph{Step 6: the engineering point.} A dimensional check on every
new equation in a draft report catches errors of substitution,
indexing, and copy-paste before they propagate to the calculation
result. The check is cheap (each side of the equation gets a
dimension expression, the dimensions are compared) and the failure
mode is sharp (a $[L]^{3}$ mismatch is unambiguous). The chapter's
discipline is to run the check before plugging in numbers; the
exercise's draft equation would have been caught at the formula
stage rather than after the calculation produced an absurd answer.
\end{exercise}

\begin{exercise}
A draft formula for the drag force on a parachute reads
\[
F = \tfrac{1}{2} \rho v A C_{D}
\]
where $\rho$ is air density, $v$ is descent speed, $A$ is parachute
area, and $C_{D}$ is the drag coefficient. State whether the formula
is dimensionally correct. If not, identify the missing factor and
explain its origin.

\paragraph{Solution.}
The drag-force formula is a working aerospace and mechanical-
engineering result. The dimensional check catches a recurring
typo (missing factor of $v$) before the calculation produces a
parachute that descends at the wrong speed.

\emph{Step 1: state the dimensions of the left-hand side.} The
force $F$ has dimensions $[F] = [M][L][T]^{-2}$.

\emph{Step 2: assemble the dimensions of the right-hand side.}
\begin{itemize}[leftmargin=*]
  \item $[\rho] = [M][L]^{-3}$ (mass per unit volume, air density).
  \item $[v] = [L][T]^{-1}$ (descent speed).
  \item $[A] = [L]^{2}$ (parachute area).
  \item $[C_{D}] = 1$ (drag coefficient is dimensionless).
  \item $[\tfrac{1}{2}] = 1$ (numerical prefactor is dimensionless).
\end{itemize}
The combined dimensions are
\[
[\rho][v][A][C_{D}] = [M][L]^{-3} \cdot [L][T]^{-1} \cdot [L]^{2}
\cdot 1 = [M][T]^{-1}.
\]

\emph{Step 3: compare to the required dimensions.} The right-hand
side has dimensions $[M][T]^{-1}$. The required dimensions are
$[M][L][T]^{-2}$. The mismatch is by $[L][T]^{-1}$, exactly one
extra power of velocity. The formula is missing a factor of $v$.

\emph{Step 4: write the corrected form.} The dimensionally correct
drag-force expression is
\[
F = \tfrac{1}{2} \rho v^{2} A C_{D}.
\]
The corrected right-hand side has dimensions $[M][L]^{-3} \cdot
[L]^{2}[T]^{-2} \cdot [L]^{2} \cdot 1 = [M][L][T]^{-2}$, matching
the left.

\emph{Step 5: identify the physical origin of the $v^{2}$.} The
factor of $v^{2}$ comes from the kinetic-energy flux into the drag
surface. The drag force per unit area is the dynamic pressure
$\tfrac{1}{2}\rho v^{2}$ (Bernoulli's working result), not the
linear-velocity expression the draft formula wrote. The $v^{2}$
scaling is why a parachute deployed at twice the descent speed
sees four times the drag force (which is why a parachute slows
descent in the first place: the increasing drag force at higher
$v$ rapidly approaches the weight, terminating the acceleration).

\emph{Step 6: the engineering point.} The dimensional check on the
draft formula takes ten seconds and is unambiguous: $[M][T]^{-1}$
is not $[F]$, off by a factor of velocity. A working aerospace
engineer would catch this before computing a single value. The
chapter's pedagogical claim is that the check is cheap and the
failure mode of skipping it is expensive: the published terminal-
velocity formula with the wrong drag expression would predict
half-the-correct descent speed, and a parachute design built to
that prediction would deploy a different size than the mission
requires.
\end{exercise}

\begin{exercise}
A spreadsheet that computes pump power from flow rate, pressure
rise, and efficiency returns $1{,}500 \,\si{\watt}$ for a pump that
the manufacturer rates at $1.5 \,\si{\kilo\watt}$. State whether
this is a unit error, a numerical coincidence, or a meaningful
disagreement, and propose the next measurement that distinguishes
the three.

\paragraph{Solution sketch.}
$1{,}500 \,\si{\watt} = 1.5 \,\si{\kilo\watt}$ exactly; the two are
the same physical quantity in different units. No disagreement
exists. This is the classic kilo-versus-no-prefix confusion: the
spreadsheet reports in watts, the manufacturer in kilowatts, and a
reader who treats them as different numbers misreads the agreement
as disagreement. The next measurement to take is no measurement at
all; the correct action is to standardise on one unit system in the
report and to confirm that the spreadsheet's output column is
labelled with its unit explicitly. If the two genuinely disagreed
(say, $1500 \,\si{\watt}$ from spreadsheet vs $2000 \,\si{\watt}$
rated), the next measurement is the actual electrical input power
under nameplate flow and head conditions.
\end{exercise}

\begin{exercise}
A datasheet specifies a capacitor as $10 \,\mathrm{mF}$ and another
as $10 \,\mathrm{\mu F}$. State the ratio of the two capacitances,
identify the typeface mistake that frequently confuses the two, and
propose a labelling rule that would catch the confusion in
hand-drawn circuit diagrams.

\paragraph{Solution sketch.}
$10 \,\mathrm{mF} = 10 \times 10^{-3} \,\si{\farad} = 10^{-2} \,\si
{\farad}$; $10 \,\mathrm{\mu F} = 10 \times 10^{-6} \,\si{\farad}
= 10^{-5} \,\si{\farad}$. Ratio: $1000\times$. The typeface mistake
that frequently confuses the two is mis-typesetting the Greek
letter $\mu$ as a Latin $u$ or a lower-case $m$; sloppy handwriting
of $\mu$ can resemble a Latin $m$. Labelling rule for hand-drawn
diagrams: spell out ``mF'' or ``microF'' or ``$10^{-6}\,$F'' rather
than rely on the prefix symbol, or use the explicit decimal form
(0.01 F and 0.000010 F). In industry practice, capacitor markings
on small SMD devices use a numerical-only code (e.g., 103 = 10
$\times$ 10$^{3}$ pF = 10 nF) to avoid the prefix-confusion class
entirely.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
The Hubble Space Telescope's primary-mirror error originated in a
reflective null corrector whose internal spacing was off by
$1.3 \,\si{\milli\meter}$. Describe (i) the measurement chain from
the metering rod to the reflective null corrector, (ii) the metrology
step that produced the wrong spacing, and (iii) the auxiliary optical
tests that indicated the error before launch. Cite the NASA Optical
Systems Failure Report \cite{acc:nasa-hubble-optical-systems-1990}.

\paragraph{Rubric.} (i) The metering rod was used to set the
longitudinal spacing of the reflective null corrector's optical
elements; the rod's end cap and an internal feature served as
reference targets; the spacing was transferred from rod to corrector
through a contact-and-readout procedure. (ii) The metrology step
that produced the wrong spacing was the misreading of the metering
rod's target: a paint-mark on the end cap was treated as the
reference, when the internal feature was the actual datum, shifting
the read by $1.3 \,\si{\milli\meter}$. (iii) Two auxiliary tests
showed discrepancies: a refractive null test (with a different
corrector) and an inverse null test. Both indicated figure error
inconsistent with the primary reflective-null-corrector
measurement. Both were set aside under cost and schedule pressure
on the grounds that the reflective corrector was the primary
reference and the others were untrusted secondary checks. The full
investigation is in the NASA OSFR \cite{acc:nasa-hubble-optical-systems-1990}.
\end{exercise}

\begin{exercise}
Therac-25 displayed terse malfunction and dose information that
operators could not interpret as a safety state. Argue, in 200 to 300
words, whether the useful classification is unit-interface failure,
software-engineering failure, human-factors failure, or regulatory
failure. State which classification produces the most useful
corrective action and why. Cite Leveson and Turner
\cite{paper:leveson-turner1993}.

\paragraph{Rubric.} A credible argument acknowledges that each
classification captures part of the failure: the dose display was a
unit-interface failure (the displayed quantity's meaning was
ambiguous to the operator); the unhandled race conditions were a
software-engineering failure; the operator workflow that habituated
``MALFUNCTION 54'' as routine was a human-factors failure; the
absence of independent safety-critical-software review and the
manufacturer's missing hazard analysis were a regulatory failure.
The most useful corrective action is the regulatory framing,
because it covers the engineering and operational layers as
subordinate practices that the regulatory regime should have
required. Subsequent regulatory action (the FDA's medical-device
software guidance; the IEC 62304 standard; the post-market
surveillance requirements that emerged in the 1990s) addressed the
underlying gap. The narrower software-engineering framing addresses
only the implementation; the unit-display framing addresses only
the operator interface; neither propagates to the systemic change
that prevents the next instance. Cite Leveson and Turner
\cite{paper:leveson-turner1993}.
\end{exercise}

\begin{exercise}
A chemical-process plant operates at $300 \,\si{\degreeCelsius}$ and
pressure $25\,\text{bar}$. A new instrument is procured whose
specification sheet rates it for $300\,{}^{\circ}\mathrm{F}$ and
$25\,\text{psi}$. State
the dimensional discipline the procurement process should have
imposed; state whether the instrument should be installed; identify
the audit step at which the discrepancy should have been caught.

\paragraph{Rubric.} The discipline: every procurement specification
carries explicit units for every rated condition, and the
purchasing engineer's review checks the specified vs the operating
units before purchase order release. The instrument should not be
installed: $300 \,{}^{\circ}\mathrm{F} \approx 149 \,\si{\degree
Celsius}$ (less than half the plant's operating temperature), and
$25 \,\text{psi} \approx 1.7 \,\text{bar}$ (less than 7\% of the
plant's operating pressure). Installation would produce immediate
overpressure and overtemperature failure with potential for
catastrophic release. The audit step at which the discrepancy
should have been caught is the technical specification review
preceding purchase order release; the procurement engineer should
have flagged the unit-system mismatch as part of the routine
review. A secondary audit step is the materials-receiving
inspection, which should match the as-received specification
against the engineering drawing's rated conditions. A tertiary
audit is the pre-commissioning test under nominal conditions, which
would reveal the failure before plant startup.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
Argue, in one paragraph, whether dimensional analysis is more useful
in fluid mechanics than in solid mechanics. Give one example from
each domain to support your position.

\paragraph{Discussion.} A credible argument acknowledges that
dimensional analysis is more visibly useful in fluid mechanics
because fluid problems frequently have many parameters and the
collapse onto dimensionless groups (Reynolds, Froude, Mach, Prandtl,
Strouhal, Bond, Grashof, Stokes) is the working language of the
discipline. The wind-tunnel scaling case is the canonical example.
In solid mechanics, dimensional analysis appears more often as a
debugging check than as a design tool, because the elastic-stress
problem is well-posed in coordinates and the relevant scaling laws
(buckling-load scaling with $EI/L^{2}$, plate-bending scaling) are
typically taught analytically rather than as dimensionless-group
collapses. An example: the buckling load of a slender column under
axial compression scales as $\pi^{2} EI / L^{2}$ (Euler's
formula); the dimensional content $[EI/L^{2}] = [F]$ is consistent,
but the engineering use is direct calculation rather than
similarity-based experimental design. The position is defensible
either way; the strongest argument acknowledges the historical
contingency of the two fields' methodological preferences.
\end{exercise}

\begin{exercise}
A senior colleague tells the reader that ``dimensional analysis is a
toy; you cannot derive a number from it.'' Write a one-paragraph
reply that agrees where agreement is honest and disagrees where
disagreement is honest, with a specific example.

\paragraph{Discussion.} A credible reply concedes the literal point:
dimensional analysis does not produce numerical prefactors. The
pendulum's $2\pi$, the drag coefficient's actual curve, the
Hagen-Poiseuille $\pi/8$, the $\gamma$ in sound-speed: dimensional
analysis cannot recover these. The reply disagrees on the broader
claim. Dimensional analysis is not a toy; it is the structural
backbone of every engineering equation. The Buckingham pi result
collapses a multi-variable parameter space onto a tractable curve,
which is what makes experimental engineering possible. The
debugging value of dimensional checks is real: a dimensional
inconsistency stops a wrong calculation before any computation
runs. The senior colleague's point is true about the prefactors;
the broader dismissal misses the structural and economic value.
The professional engineer uses dimensional analysis as both
debugging tool and design framework, and saves the prefactor for
the differential equation or the experiment that earns it.
\end{exercise}
