\chapter{Tissue and organ engineering}
\label{vol06:ch05}

\epigraph{The cell is the smallest formal unit of life, but the
organism is not a cell: it is an architecture of cells, of matrices,
of vessels, of nerves, and of the negotiated equilibria among them.}{Y.\ C.\
Fung, \emph{Biomechanics: Mechanical Properties of Living Tissues}
(2nd ed., 1993), preface \cite[p.~ix]{text:fung-biomechanics}.}

\chapmeta{Half-life: foundational core (tissue mechanics, vascularisation constraints) is durable; specific scaffold materials and 3D-printing protocols are medium-life and tagged accordingly. Archetypes: scaling, interface. Exercise target: 25. Project track: design study (vascularised scaffold geometry, defended in writing). Hazard class: none (paper exercise). Reader path default: standard, with sections explicitly tagged. Named cases: Macchiarini synthetic-trachea transplants 2008-2014 (\texttt{acc:macchiarini-trachea-2008-2014}).}

\noindent
The chapter scales from the cell to the organism. A single cell is
the unit; a tissue is a population of cells organised into a working
collective; an organ is a collection of tissues organised to perform
a body-scale function; an organ system is a network of organs that
collectively maintains a physiological variable. Engineering the
intermediate scales is the subject of tissue engineering. We develop
the hierarchy, the extracellular matrix that holds it together, the
mechanics of tissues, the vascularisation constraint that limits all
tissue construction, the scaffold-and-print methods that the field
uses, the stem-cell and organoid technologies that supply the
cellular material, and the failure modes that haunt the translational
clinic. The chapter establishes its working numbers in two estimation
blocks (vasculature volume fraction, capillary total length) and
tests its constitutive thinking on the Macchiarini synthetic-trachea
case (2008-2014), the canonical clinical-scale failure of synthetic-
scaffold whole-organ engineering.

\section{The hierarchy: cells, tissues, organs, systems}\pathtag{core}

\engterm{Tissue}. A tissue is a population of cells, generally of one
or a few types, organised in a characteristic pattern and embedded in
an extracellular matrix. Histologists divide tissues into four
classical types: \engterm{epithelial} (sheets of cells covering
surfaces and lining cavities, with apical-basal polarity), \engterm{connective}
(cells dispersed in a substantial extracellular matrix, including
bone, cartilage, tendon, ligament, fat, and the loose connective
tissue between organs), \engterm{muscle} (contractile cells:
skeletal, cardiac, smooth), and \engterm{nervous} (neurons and glial
support cells). The four-type classification dates from the
nineteenth-century histologists and remains the working vocabulary;
modern tissue engineering adds finer subclassifications based on
cell lineage and developmental origin.

\engterm{Organ}. An organ is a structurally bounded assembly of
tissues that performs a defined body-scale function. The human heart
is composed of cardiac muscle (the bulk), epithelial endocardium
(the inner lining), epicardial fibroblasts and adipocytes (the outer
surface), a coronary vasculature (vessels providing oxygen to the
heart muscle itself), and an electrical conduction network of
specialised cardiomyocytes (the sinoatrial node, atrioventricular
node, His-Purkinje system). Engineering an organ-equivalent in vitro
requires reproducing not just the dominant cell type but the entire
tissue mosaic in the appropriate geometry.

\engterm{System}. An organ system is a coordinated set of organs that
maintains a physiological function. The cardiovascular system
(heart, arteries, veins, capillaries) maintains tissue perfusion;
the respiratory system (lungs, airways, diaphragm) maintains gas
exchange; the digestive system (mouth, stomach, intestines, liver,
pancreas) processes food and absorbs nutrients; the urinary system
(kidneys, ureters, bladder, urethra) maintains fluid and electrolyte
balance; the nervous system (brain, spinal cord, peripheral nerves)
coordinates the others (Chapter~\ref{vol06:ch07}). System-level
engineering (organ-on-chip platforms, body-on-chip systems) attempts
to recapitulate inter-organ interactions in miniature culture.

\engterm{Reading the hierarchy as scaling}. The transition from cell
to system spans seven orders of magnitude in length scale (a
$10\,\si{\micro\meter}$ cell to a $1\,\text{m}$ adult human) and
twelve orders of magnitude in mass (a $10^{-12}\,\text{g}$ cell to a
$70\,\text{kg}$ adult). The engineering constraints change at each
scale. Cells are limited by diffusion across the plasma membrane;
tissues by diffusion across the tissue thickness; organs by the
geometry of their internal transport network (vasculature); systems
by inter-organ feedback loops. The volume's scaling archetype
appears in its canonical biological form in this hierarchy: each
scale solves the scaling-up problem of the previous scale by
inventing a new transport mechanism. Figure~\ref{fig:vol06ch05:hierarchy}
shows the four levels with the dominant transport mechanism at each.

\input{volumes/06-life/ch05-tissue-and-organ-engineering/figures/fig-tissue-hierarchy.tex}

\engterm{Worked numbers}. The numerical span of the hierarchy is
worth setting before we proceed. A typical mammalian cell has a
diameter of order $10\,\si{\micro\meter}$, a volume of order
$5 \times 10^{-16}\,\text{m}^3$, and a mass of order
$5 \times 10^{-13}\,\text{kg}$. The adult human body
($\sim 70\,\text{kg}$) contains of order $3 \times 10^{13}$ cells.
A typical organ ($\sim 100\,\text{g}$ to $\sim 1\,\text{kg}$)
contains of order $10^{11}$ cells. The capillary network of the
whole body has a total length of $50\,000$ to $80\,000\,\text{km}$;
the second estimation block of this chapter derives the working number
from blood volume and capillary radius rather than direct
measurement.

\engterm{The scaling-of-transport ladder}. Across the seven decades
of length, four distinct transport mechanisms operate. (i)~Within a
cell, intracellular transport happens by diffusion and by active
cytoskeletal transport (motor proteins on microtubules and actin),
with characteristic diffusion times $\tau = L^2/D$ of milliseconds
for $L \sim 10\,\si{\micro\meter}$ and $D \sim
10^{-10}\,\text{m}^2/\text{s}$ for a small molecule. (ii)~Within a
tissue, extracellular diffusion supplies cells up to the diffusion
limit ($L \sim 100$--$200\,\si{\micro\meter}$ for oxygen). (iii)~Within
an organ, bulk flow through the vasculature carries the cells'
nutrients to and from the diffusion-supplied tissue layer. (iv)~Within
the body, the inter-organ transport network (the cardiovascular
system) plus the endocrine system maintains the inter-organ
homeostatic equilibria. The engineer constructing a tissue substitute
inherits the diffusion limit as the binding constraint and must
provide the bulk-flow infrastructure that nature installs in the
form of the vasculature.

\section{The extracellular matrix}\pathtag{standard}

The \engterm{extracellular matrix} (ECM) is the protein-and-polysaccharide
gel that fills the space between cells and gives tissues their
mechanical, biochemical, and signalling properties. The ECM is
synthesised by the tissue's cells and remodelled continuously by the
combination of synthesis, secretion, and degradation by specific
enzyme families (the matrix metalloproteinases).

\engterm{Principal components}. The ECM contains five principal
classes of macromolecule (Figure~\ref{fig:vol06ch05:ecm}).
\engterm{Collagens} (a family of $\sim 28$
homologous proteins, of which type I, II, III, and IV are the most
abundant) form the load-bearing fibrils of connective tissue. Type-I
collagen in bone, tendon, and skin has a tensile strength of
approximately $100\,\text{MPa}$ and a Young's modulus of $1$ to
$2\,\text{GPa}$ along the fibre direction. \engterm{Elastin} forms
the elastic fibres of skin, lung, and arterial walls; elastin can
stretch reversibly to $130\,\%$ of its resting length. \engterm{Proteoglycans}
(a core protein with attached glycosaminoglycan chains) bind water
and provide compressive resistance; aggrecan in cartilage is the
canonical example, with chondroitin sulfate side chains that hold
water at $80\,\%$ of cartilage's mass. \engterm{Glycoproteins}
(fibronectin, laminin, vitronectin) mediate cell-matrix attachment
through specific cell-surface receptors (integrins). \engterm{Hyaluronic
acid} is a large unbranched polysaccharide that contributes to the
viscoelasticity of soft tissues.

\input{volumes/06-life/ch05-tissue-and-organ-engineering/figures/fig-ecm-components.tex}

\engterm{Specialised matrices}. Bone matrix contains collagen
($\sim 30\,\%$ by mass) plus the mineral hydroxyapatite
(Ca$_5$(PO$_4$)$_3$(OH), $\sim 60\,\%$ by mass) plus water. The
collagen provides toughness; the mineral provides stiffness; the
composite has properties superior to either component alone. Tendon
matrix is approximately $70\,\%$ collagen by dry mass, oriented in
parallel along the load direction. Cartilage matrix has high
proteoglycan content and low cell density, with a Young's modulus
of $0.5$ to $1\,\text{MPa}$ and a friction coefficient below
$0.01$ when lubricated. Skin matrix is a mixture of collagen,
elastin, and ground substance with regional variation in composition
that maps onto regional differences in mechanical demands.

\engterm{Matrix as instruction substrate}. The ECM is not only a
structural filler. It is an active instruction substrate for the
cells embedded in it. Matrix stiffness directly influences cell fate:
mesenchymal stem cells cultured on substrates of stiffness comparable
to brain tissue ($\sim 1\,\text{kPa}$) differentiate toward neurons,
those on muscle-like stiffness ($\sim 10\,\text{kPa}$) toward
myocytes, those on bone-like stiffness ($\sim 30\,\text{kPa}$) toward
osteoblasts \cite{paper:v6c05-engler-stiffness-2006}. The 2006
discovery by Engler, Sen, Sweeney, and Discher was one of the most
consequential in cell biology and has driven a decade of work on
mechanotransduction \cite{paper:v6c05-discher-mechanotransduction-2009,
paper:v6c05-peyton-stiffness-review-2007}. Matrix composition
similarly signals to cells through integrin-mediated attachment and
through the local concentration of bound growth factors. The engineer
designing a scaffold for tissue regeneration must specify the
matrix's mechanical properties, composition, and bound-signal
profile, not just its geometry
\cite[ch.~24]{text:v6c05-langer-vacanti-tissue-2014}.

\engterm{Worked example: matrix stiffness sets cell fate}. The
Engler 2006 paper deposited human mesenchymal stem cells on
polyacrylamide gels of three stiffnesses: $0.1$--$1\,\text{kPa}$
(brain-like), $8$--$17\,\text{kPa}$ (muscle-like), and
$25$--$40\,\text{kPa}$ (osteoid-like). The cells were cultured in
identical media for one week. The stiffness alone, with no soluble
fate-determining factor, drove the cells toward neuronal markers
($\beta$3-tubulin, $\sim 60\,\%$ of cells, on the soft gel), myogenic
markers (MyoD, $\sim 40\,\%$ of cells, on the intermediate gel), and
osteogenic markers (CBF$\alpha$1, $\sim 50\,\%$ of cells, on the
stiff gel). The fate commitment was lineage-specific and not the
result of differential cell adhesion or proliferation, ruled out by
controls. The mechanism, subsequently traced, involves
acto-myosin-generated tension on integrin-mediated focal adhesions;
the cell senses the substrate stiffness by pulling on it and reading
the resistance. The scaffold engineer's working consequence: every
specification of a scaffold must include the matrix stiffness in the
range relevant to the target tissue, with order-of-magnitude
precision, because off-target stiffness will divert cell fate. A
soft scaffold seeded with bone-marrow-derived mesenchymal stem cells
will produce neuron-like cells regardless of the developer's
intention. The Engler result is the foundation of the
``stiffness-matched scaffold'' principle that all credible scaffold
designs since 2006 have respected.

\engterm{Matrix mechanics worked example: a $5\,\text{mm}$ tendon
strip in uniaxial tension}. A common preclinical test is a $5\,\text{mm}$
wide, $20\,\text{mm}$ long, $1\,\text{mm}$ thick strip of tendon
tested in uniaxial tension. With a fibre-aligned Young's modulus of
$1\,\text{GPa}$, a $5\,\text{mm} \times 1\,\text{mm}$ cross-section
loaded to $1\,\%$ strain ($\Delta L = 200\,\si{\micro\meter}$) carries
\[
F = E \cdot A \cdot \varepsilon
  = 10^{9}\,\text{Pa} \times 5 \times 10^{-6}\,\text{m}^2 \times 0.01
  = 50\,\text{N}.
\]
The same strip at the failure strain of $\sim 10\,\%$ carries
$\sim 500\,\text{N}$ at the linear extrapolation, more typically
$\sim 250\,\text{N}$ once strain-stiffening departures from
linearity are accounted for. A tissue-engineered tendon replacement
that must carry physiological loads ($\sim 50$--$100\,\text{N}$ for
the supraspinatus, $\sim 1000$--$3000\,\text{N}$ for the Achilles)
must reach modulus and ultimate-strength figures of this order at
the time of implantation, not after months of in-vivo remodelling.
This is the constraint that has limited tissue-engineered tendon
constructs to a research-grade product since the field's foundation.

\section{Tissue mechanics}\pathtag{core}

Tissues are mechanical objects with stress-strain relations that the
engineer must measure and model. The relations are usually non-linear
(stress-strain curves are not Hookean for most tissues), anisotropic
(properties differ along different directions), viscoelastic (the
response depends on the rate of loading), and history-dependent
(prior loading alters the current response). The mechanics of soft
tissues are the subject of Y.~C. Fung's foundational monograph
\cite{text:fung-biomechanics} and Chapter~\ref{vol06:ch06} will
return to them in the context of biomechanics; the present chapter
develops the constitutive-modelling framework that the engineering
of scaffolds and synthetic tissues requires.

\engterm{Stress and strain}. We use the standard continuum-mechanics
definitions. \engterm{Cauchy stress} $\sigma$ is force per unit
current area; \engterm{Green strain} $E$ is a finite-strain measure
appropriate to large deformations. For small deformations
(strain $\le 5\,\%$), the linearised infinitesimal-strain framework
is adequate; for the larger deformations typical of soft tissues
($> 10\,\%$), finite-strain measures are necessary.

\engterm{Hyperelastic constitutive models}. A common framework for
soft tissues is hyperelasticity: the stress is derived from a
strain-energy density function $W(E)$ such that $\sigma_{ij} =
\partial W/\partial E_{ij}$. For an isotropic tissue, $W$ depends
only on the strain invariants $I_1$, $I_2$, $I_3$. Common forms
include the neo-Hookean model
$W = \frac{\mu}{2}(I_1 - 3)$, the Mooney-Rivlin model
$W = c_1(I_1 - 3) + c_2(I_2 - 3)$, and the Fung exponential form
$W = \frac{1}{2} c (e^{Q} - 1)$ with $Q$ a quadratic form in the
strain components. The Fung model fits a wide range of soft tissues
because the exponential captures the strain-stiffening behaviour
typical of collagen-containing tissues
\cite[ch.~7]{text:fung-biomechanics}. Each model has two to six fitting
parameters that the engineer measures by uniaxial or biaxial tension
tests on the tissue of interest.

\engterm{Viscoelasticity}. Soft tissues dissipate energy when
deformed; the dissipation appears as hysteresis in the loading-unloading
cycle. The standard viscoelastic models (Kelvin-Voigt, Maxwell,
standard linear solid) combine springs and dashpots in different
topologies. The relaxation modulus $G(t)$ of a typical soft tissue
follows a power law $G(t) \sim t^{-\alpha}$ with $\alpha \approx 0.1$
to $0.3$ over many decades in time. The quasi-linear viscoelastic
(QLV) framework of Fung \cite{paper:v6c05-fung-qlv-1972}, with a
strain-dependent elastic response and a strain-independent
relaxation function, captures most of the phenomenology.

\engterm{Anisotropy}. Tissues organised by fibres (tendon, ligament,
arterial wall, cardiac muscle) are mechanically anisotropic. The
constitutive models for anisotropic tissues include the
Holzapfel-Gasser-Ogden (HGO) family
\cite{paper:v6c05-holzapfel-hgo-2000}, which adds collagen fibre
contributions to a neo-Hookean ground substance. The HGO model and its
variants are the working standard for arterial mechanics and have
been incorporated into commercial finite-element codes.

\input{volumes/06-life/ch05-tissue-and-organ-engineering/figures/fig-tissue-stress-strain.tex}

\engterm{Tissue-specific properties}. The principal soft tissues have
characteristic mechanical properties at small strain: brain
$\sim 1\,\text{kPa}$, fat $\sim 5\,\text{kPa}$, muscle (passive)
$\sim 10\,\text{kPa}$, liver $\sim 1$ to $5\,\text{kPa}$, skin
$\sim 100\,\text{kPa}$, cartilage $\sim 500\,\text{kPa}$ to
$1\,\text{MPa}$, tendon $\sim 1\,\text{GPa}$ along the fibre,
cortical bone $\sim 15\,\text{GPa}$. The five orders of magnitude
span reflect the structural composition: protein-rich, water-rich
tissues are soft; collagen-rich tissues are stiff; mineralised
tissues are stiffer still. The full reference table is at
\path{data/tissue_mechanical_properties.csv} and is the dataset for
the simulation and design exercises.

\engterm{Worked example: Mooney-Rivlin parameters from a uniaxial
test}. Suppose a soft-tissue specimen is tested in uniaxial tension
and produces a stress-strain curve with apparent low-strain modulus
$E_0 = 200\,\text{kPa}$ at $\varepsilon \to 0$ and a strain-stiffening
factor of $3$ (the secant modulus at $\varepsilon = 0.5$ equals
$3\,E_0$). The incompressible Mooney-Rivlin Cauchy stress in
uniaxial tension is
\[
\sigma(\lambda) = 2\!\left(c_1 + \frac{c_2}{\lambda}\right)
                  \!\left(\lambda^2 - \frac{1}{\lambda}\right),
\]
with $\lambda = 1 + \varepsilon$ the stretch ratio. At $\lambda \to 1$
the small-strain modulus is $E_0 = 6(c_1 + c_2)$, fixing
$c_1 + c_2 = 33\,\text{kPa}$. The secant-modulus condition at
$\lambda = 1.5$ gives a second equation in $c_1, c_2$; solving the
two simultaneously yields representative values $c_1 \approx
20\,\text{kPa}$, $c_2 \approx 13\,\text{kPa}$. Both must be
non-negative (Drucker stability), which constrains the achievable
stiffening factor: Mooney-Rivlin captures stiffening only up to a
factor of $\sim 5$ before the form becomes inadequate, at which
point the engineer switches to the Fung exponential. The fit and the
limits of fit are implemented in
\path{code/tissue_stress_strain_fit.py}.

\begin{principle}[Tissue mechanics is constitutive, not Hookean]
The mechanical behaviour of a soft tissue is a non-linear, anisotropic,
viscoelastic constitutive relation that the engineer measures and
fits for each tissue type. The Hookean approximation applies only at
small strain; the engineering models (Fung, Mooney-Rivlin, HGO) are
the working standard for tissue mechanics. A scaffold whose stiffness
matches the target tissue in the small-strain regime may still fail
the strain-stiffening or anisotropy requirements at physiological
loads. The engineer specifies the constitutive model, not just one
modulus value.
\end{principle}

\section{Vascularisation as design constraint}\pathtag{core}

The single hardest engineering problem in tissue construction is
vascularisation. A piece of tissue thicker than the diffusion limit
($\sim 200\,\si{\micro\meter}$ for oxygen delivery in metabolically
active tissue) cannot survive without an internal blood supply. The
vasculature must extend through the tissue with sufficient density
that no cell is more than $\sim 100\,\si{\micro\meter}$ from a
capillary, must connect to the host's circulation upon implantation,
must withstand the blood pressure ($\sim 100\,\text{mmHg}$ peak
systolic in the arterial tree), and must maintain anti-thrombotic
function over the lifetime of the implant.

\engterm{The diffusion limit}. The maximum tissue thickness that can
survive on diffusion from one face is set by the steady-state
oxygen-diffusion equation. The oxygen consumption rate per unit
volume in metabolically active tissue is approximately
$\dot{q} = 10^{-3}\,\text{mol}/(\text{m}^3\cdot\text{s})$. The
oxygen diffusion coefficient in tissue is $D \approx 2 \times 10^{-9}\,
\text{m}^2/\text{s}$. The arterial oxygen partial pressure is
$\sim 100\,\text{mmHg}$, corresponding to a dissolved-oxygen
concentration of $C_0 \approx 0.13\,\text{mM}$ in plasma.

\engterm{Krogh cylinder derivation}. The cleanest derivation runs
on the cylindrical geometry that August Krogh introduced in 1919
to read capillary-supplied muscle \cite{paper:v6c05-krogh-1919}.
A capillary of radius $R_c \approx 4\,\si{\micro\meter}$ at the
centre of a tissue cylinder of radius $R_t$ supplies oxygen by radial
diffusion (Figure~\ref{fig:vol06ch05:krogh}). At steady state, the
oxygen-balance equation in cylindrical coordinates is
\[
\frac{1}{r}\frac{\dd}{\dd r}\!\left(r \frac{\dd C}{\dd r}\right)
  = \frac{\dot{q}}{D},
\]
with boundary conditions $C(R_c) = C_0$ (capillary surface) and
$\dd C/\dd r |_{R_t} = 0$ (symmetry on the cylinder boundary).
Integrating once:
\[
r\frac{\dd C}{\dd r} = \frac{\dot{q}}{2D}\,r^2 + A,
\]
and applying the symmetry condition at $r = R_t$ gives
$A = -\dot{q} R_t^2/(2D)$. Integrating again and applying the
capillary boundary condition:
\[
C(r) = C_0 + \frac{\dot{q}}{4D}(r^2 - R_c^2)
       - \frac{\dot{q} R_t^2}{2D}\ln\!\left(\frac{r}{R_c}\right).
\]
The maximum tissue radius the capillary can supply is the $R_t$
at which $C(R_t) = 0$:
\[
0 = C_0 + \frac{\dot{q}}{4D}(R_t^2 - R_c^2)
       - \frac{\dot{q} R_t^2}{2D}\ln\!\left(\frac{R_t}{R_c}\right).
\]
Solving numerically with the working numbers above (and
$R_c = 4\,\si{\micro\meter}$) gives $R_t \approx 50$ to
$70\,\si{\micro\meter}$, depending on consumption rate. This is the
Krogh number; it sets the inter-capillary spacing in tissue at
$\sim 100$--$200\,\si{\micro\meter}$ centre to centre.

\input{volumes/06-life/ch05-tissue-and-organ-engineering/figures/fig-krogh-cylinder.tex}

\engterm{Slab simplification}. For a slab of tissue of thickness $L$
supplied from one face (e.g., a thin-construct skin substitute),
the same balance gives the closed-form maximum thickness
\[
L_{\max} = \sqrt{\frac{2 D C_0}{\dot{q}}}
        \approx \sqrt{\frac{2 \times 2\times 10^{-9} \times 0.13}{10^{-3}}}
        \,\text{m}
        \approx 230\,\si{\micro\meter}.
\]
The slab and Krogh-cylinder estimates agree within a factor of two;
the working number $L \sim 100$--$200\,\si{\micro\meter}$ is the
engineering diffusion limit. The numerical sensitivity to the input
parameters is implemented at
\path{code/krogh_oxygen_diffusion.py}; the sweep over $D$ and
$\dot{q}$ shows that the limit scales as $\sqrt{D/\dot{q}}$,
moving by a factor of three across the physiological range of
either parameter.

\engterm{Strategies}. Tissue engineering uses four principal
strategies to address the vascularisation constraint
\cite{paper:v6c05-novosel-vascularisation-review-2011}. \engterm{Thin
constructs} (skin substitutes, corneal grafts) are below the
diffusion limit and do not require internal vasculature.
\engterm{Pre-vascularised constructs} have a microvascular network
seeded into the scaffold in vitro, with the network anastomosing
(connecting) to host vessels after implantation; the iPSC-derived
liver bud of Takebe et al.\ 2013 \cite{paper:v6c05-takebe-liver-2013}
is an early proof of concept. \engterm{Channel-based constructs}
have pre-formed channels through the scaffold that act as artificial
vessels; the host endothelialises the channels after implantation
(Miller et al.\ 2012 \cite{paper:v6c05-miller-vascular-print-2012}
introduced the sacrificial-template approach).
\engterm{Sequential implantation} grows the construct in a
pre-vascularised host site first (a flap or chamber with its own
blood supply) and only then transfers it to the final location; this
strategy is used in several experimental whole-organ approaches.

\engterm{Vascular tree geometry}. The branching ratio and diameter
scaling of a vascular tree are constrained by Murray's law
\cite{paper:v6c05-murray-1926}: at every bifurcation, the cube of
the parent vessel diameter equals the sum of the cubes of the
daughter diameters ($d_{\text{parent}}^3 = d_1^3 + d_2^3$). The law
minimises the metabolic cost of maintaining a blood vessel (which
scales with vessel volume) plus the work cost of pumping blood
(which scales with the inverse of $d^4$ by Poiseuille's law).
Murray's law fits the observed branching of mammalian arterial trees
to within $\sim 10\,\%$ across many orders of magnitude in vessel
size and is the engineering basis of scaffold-design strategies that
try to recapitulate natural vasculature
\cite[ch.~5]{text:phillips-physical-biology}. For a symmetric
bifurcation, the daughter-to-parent diameter ratio is
$2^{-1/3} \approx 0.794$; for the optimal symmetric bifurcation angle,
minimising the same cost yields $\theta \approx 37.5^{\circ}$.
Figure~\ref{fig:vol06ch05:murray} shows the geometry; the
hierarchy from a $4\,\text{mm}$ root through eight generations is
tabulated by \path{code/murrays_law_branching.py}.

\input{volumes/06-life/ch05-tissue-and-organ-engineering/figures/fig-murray-bifurcation.tex}

\engterm{Capillary density}. The human body contains approximately
$10^{10}$ capillaries with a total length of $50\,000$ to $80\,000\,\text{km}$
and a total exchange surface area of $\sim 100\,\text{m}^2$. The
capillary density is tissue-specific: $2000$ to $3000\,\text{mm}^{-2}$
in cardiac muscle and active skeletal muscle, $300$ to
$500\,\text{mm}^{-2}$ in resting muscle, $200$ to
$1000\,\text{mm}^{-2}$ in nervous tissue, and lower in connective
tissue. The reader designing a vascularised scaffold targets the
capillary density of the tissue being replaced.

\begin{estimation}
\textbf{Question.} Estimate the volume of vasculature required to
adequately perfuse a $5\,\text{cm} \times 5\,\text{cm} \times 5\,\text{cm}$
engineered liver tissue construct.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The construct volume is $125\,\text{cm}^3$. Native liver has a
capillary density of approximately $500$ to $1000\,\text{mm}^{-2}$
when sectioned (capillaries per square millimetre of tissue
cross-section).

A $1\,\text{mm}^2$ cross-section with $500$ capillaries each of
diameter $8\,\si{\micro\meter}$ has a capillary volume per
millimetre depth of $500 \times \pi \times (4 \times 10^{-3}\,\text{mm})^2
\times 1\,\text{mm} = 2.5 \times 10^{-2}\,\text{mm}^3$. The
capillary volume fraction is $2.5\,\%$ of the tissue volume.

Adding the larger vessels (the construct must be supplied by
arterioles and venules), which collectively constitute another
$5\,\%$ to $10\,\%$ of tissue volume, gives a total vasculature
fraction of approximately $10\,\%$.

For a $125\,\text{cm}^3$ construct, the vasculature volume is
$\sim 12.5\,\text{cm}^3$. With an additional $1$ to $2\,\text{cm}^3$
for the inlet and outlet vessels, the total vasculature requirement
is approximately $15\,\text{cm}^3$ or $12\,\%$ of the construct.

\smallskip
\textit{Postmortem.} The estimate is consistent with the observed
vasculature volume fraction in native liver tissue ($\sim 10\,\%$,
including hepatic sinusoids and the larger portal-venous and arterial
vessels). The number is large: a tenth of the construct's volume must
be allocated to non-cellular vascular space. The implication is that
engineering a fully cellular tissue construct of liver volume is not
possible without internal vasculature, and that any scaffold design
must reserve $10\,\%$ of the volume for vessels before the cellular
component can be sized. The estimate is the working number for
scaffold design budgets.
\end{estimation}

\begin{estimation}
\textbf{Question.} Estimate the total length of the human capillary
network from the blood volume and the typical capillary radius.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The human blood volume is $V_b \approx 5\,\text{L} = 5 \times
10^{-3}\,\text{m}^3$. The blood is partitioned among the chambers
of the cardiovascular system in well-tabulated proportions: roughly
$\sim 65\,\%$ in the venous system (including the systemic and
pulmonary veins, the largest single reservoir), $\sim 13\,\%$ in the
pulmonary circulation as a whole, $\sim 7\,\%$ in the heart chambers,
$\sim 15\,\%$ in the systemic arteries and arterioles, and only
$\sim 5\,\%$ in the systemic capillaries at any instant. Capillary
blood volume is therefore $V_c \approx 0.05 \times 5\,\text{L}
= 250\,\text{mL} = 2.5 \times 10^{-4}\,\text{m}^3$.

Capillary radius is $R_c \approx 4\,\si{\micro\meter}$, giving
cross-sectional area $A_c = \pi R_c^2 = 5 \times 10^{-11}\,\text{m}^2$.
The total capillary length is
\[
L_{\text{cap}} = \frac{V_c}{A_c}
              = \frac{2.5 \times 10^{-4}}{5 \times 10^{-11}}\,\text{m}
              = 5 \times 10^{6}\,\text{m}
              = 5000\,\text{km}.
\]

\smallskip
\textit{Postmortem.} The published working number for the total
human capillary length is $\sim 50\,000$\,--\,$80\,000\,\text{km}$,
an order of magnitude larger than the back-of-envelope estimate
above. The discrepancy traces to the assumed capillary-blood
fraction. Physiology textbooks routinely quote $\sim 5\,\%$ as the
instantaneous fraction in the resting state, but the geometric total
length is computed from the much larger fraction that the capillary
bed could hold if fully open. Approximately $80\,\%$ of capillaries
at rest are temporarily closed (capillary recruitment); the open
fraction varies tissue-by-tissue and minute-by-minute. Correcting
for capillary recruitment ($5\,\% \to 50\,\%$ open volume in the
geometric sense) brings the estimate to $\sim 50\,000\,\text{km}$,
in line with the published number. The lesson: vasculature
measurements that conflate ``instantaneous content'' with
``anatomical lumen volume'' differ by an order of magnitude. The
engineer specifying a vascularised scaffold uses the geometric
length (capillary bed at the open state) for the perfusion budget,
because the construct's instantaneous flow capacity must accommodate
the recruitment-on-demand state, not the resting baseline.
\end{estimation}

\section{Scaffolds and 3D-printed tissue}\pathtag{standard}

A \engterm{scaffold} is a three-dimensional structure that supports
the attachment, growth, and organisation of cells into a tissue. The
scaffold provides the geometric template, the mechanical environment,
and the initial signalling context; the cells perform the eventual
tissue construction.

\engterm{Scaffold materials}. Scaffolds are made from a wide range of
materials, each with its trade-offs (the working dataset is at
\path{data/scaffold_materials.csv}). \engterm{Natural polymers}
(collagen, fibrin, hyaluronic acid, alginate) provide native cell-binding
signals and degrade through enzymatic remodelling; their batch-to-batch
variability is the main drawback for translational use.
\engterm{Synthetic polymers} (polylactic acid PLA, polyglycolic acid
PGA, their copolymer PLGA, polycaprolactone PCL) offer reproducible
properties and tunable degradation rates of weeks to years; their
drawback is the absence of native binding signals. \engterm{Ceramics}
(hydroxyapatite, tricalcium phosphate, bioactive glass) are used for
bone scaffolds because they bond directly to mineralising bone.
\engterm{Decellularised tissue} (the native ECM of a tissue with its
cells removed by detergent treatment) provides the most accurate
physiological template but requires donor tissue and adds
immunological complexity. The seminal demonstration of perfusion
decellularisation for a whole organ is Ott et al.\ 2008's
decellularised rat heart, re-seeded and producing measurable
contractile function in a bioreactor \cite{paper:v6c05-ott-decellularised-heart-2008}.

\engterm{Pore architecture}. The scaffold's pore size and
connectivity govern cell infiltration and vascular ingrowth.
Figure~\ref{fig:vol06ch05:scaffold} contrasts the closed-cell,
open-cell, and fully interconnected architectures. The design window
depends on the target tissue: $60$ to $200\,\si{\micro\meter}$ for
dermal regeneration \cite{paper:v6c05-yannas-burn-1980},
$100$ to $400\,\si{\micro\meter}$ for bone ingrowth,
$50$ to $300\,\si{\micro\meter}$ for vascular ingrowth, and
$5$ to $50\,\si{\micro\meter}$ for nerve guidance. Pore-size
distribution is process-dependent: freeze-drying produces pore-size
distributions whose median scales with freezing rate
\cite{paper:v6c05-pruitt-pore-size-1994}; electrospinning produces
sub-micrometre fibre diameters and a fibre-pore distribution that
depends on spinneret voltage and solvent viscosity.

\input{volumes/06-life/ch05-tissue-and-organ-engineering/figures/fig-scaffold-porosity.tex}

\engterm{Fabrication methods}. Scaffolds are fabricated by a portfolio
of methods. \engterm{Foaming and freeze-drying} produce porous
sponges with random pore architecture; pore sizes are typically
$100$ to $500\,\si{\micro\meter}$, suitable for cell infiltration.
\engterm{Electrospinning} produces non-woven mats of submicrometre
fibres that mimic the topology of native ECM. \engterm{3D printing}
(extrusion-based, ink-jet, stereolithography, two-photon polymerisation)
produces patient-specific geometries with sub-100-micrometre resolution
\cite{paper:v6c05-billiet-3d-printing-2012}. \engterm{Bioprinting}
extrudes cell-laden hydrogels directly, producing constructs that are
cell-seeded from the start. The choice of method depends on the
required geometric precision, the required cell density, and the
scale of the construct.

\engterm{The state of the field}. Tissue engineering at the time of
writing (current as of 2025) has produced clinical-grade products in
several areas: skin substitutes (Apligraf, Dermagraft), cartilage
implants (autologous chondrocyte implantation, FDA-approved since
1997), bone-void fillers, and corneal grafts. Solid organs (kidneys,
livers, hearts, lungs) remain in pre-clinical development; the
vascularisation barrier has limited progress for the past three
decades and remains the dominant engineering bottleneck
\cite[ch.~30]{text:v6c05-langer-vacanti-tissue-2014,
paper:v6c05-griffith-naughton-2002}. The line between the two
categories is the cellular thickness: thin constructs (millimetre
order) translate; thick constructs (centimetre order) do not, with a
short list of mainly soft-tissue exceptions. The reader pricing the
clinical near-term should expect new variants of the established
thin-construct categories (e.g., wound-bed dressings, drug-eluting
patches, autologous chondrocyte products) rather than novel
solid-organ replacements, on a five-year horizon.

\engterm{The Macchiarini case as a cautionary precedent}. Between
2008 and 2014, the thoracic surgeon Paolo Macchiarini and
collaborators implanted nine engineered tracheae in human patients:
the 2008 Lancet decellularised-cadaveric-scaffold case in Barcelona
\cite{paper:v6c05-macchiarini-2008-trachea}, then eight subsequent
implantations of synthetic-polymer (polyethylene terephthalate /
polyurethane) scaffolds at Karolinska, Krasnodar, and Peoria
between 2011 and 2014. The synthetic-scaffold programme rested on
the hypothesis that bone-marrow mononuclear cells seeded for
forty-eight hours in a bioreactor would, after implantation,
recruit host vasculature, regenerate respiratory epithelium, and
remodel the polymer into a viable airway. The hypothesis had no
animal-model foundation \cite{paper:v6c05-delaere-2014-thorax,
paper:v6c05-karolinska-heckscher-2018}. Seven of the eight
synthetic-scaffold recipients died from scaffold-related
complications, and the supporting publications were subsequently
retracted after the 2018 Heckscher external review found
misrepresentation of preclinical evidence, post-operative course,
and consent \cite{paper:v6c05-jungebluth-2011-lancet-retraction,
paper:v6c05-vogel-2017-science-investigation,
acc:macchiarini-trachea-2008-2014}. The case is the canonical
failure of synthetic-scaffold whole-organ engineering at clinical
scale; the failure section returns to its technical mechanism. The
intermediate moral for the scaffold-engineering reader is that
extrapolation from a single high-profile decellularised-tissue case
(however reported) to a synthetic-polymer programme on a different
material system without intermediate animal-model validation is the
mistake; the patient harm is the consequence. The early translational
field has at least one positive precedent, the Atala
tissue-engineered autologous bladders of 2006
\cite{paper:v6c05-atala-bladder-2006}, in which seven paediatric
patients received bladders engineered from their own urothelial
and smooth-muscle cells on biodegradable composite scaffolds and
were followed up to forty-six months with improved bladder
compliance. The Atala case is the contrasting positive demonstration:
small-volume, hollow-organ, autologous cell source, biodegradable
scaffold, with intermediate animal validation. The Macchiarini case
is the contrasting cautionary one: comparable hollow-organ geometry,
much larger construct (a $12\,\text{cm}$ airway segment with
$2\,\text{mm}$ thick walls), synthetic-polymer scaffold,
forty-eight-hour seeding rather than weeks of in-vitro tissue
formation, no animal-model intermediate. The engineering chain that
distinguishes the two cases is exactly the one the scaffold designer
controls.

\section{Stem cells and organoids}\pathtag{standard}

A \engterm{stem cell} is a cell that can both self-renew (produce
more stem cells) and differentiate into one or more specialised cell
types. Stem cells supply the cellular material for tissue engineering
and the regenerative capacity of adult tissues.

\engterm{Stem-cell hierarchy}. The hierarchy of stem-cell potency
spans four levels. \engterm{Totipotent} cells (the early embryo, up
to about eight cells) can give rise to a complete organism plus the
extra-embryonic tissues. \engterm{Pluripotent} cells (the inner cell
mass of the blastocyst, plus laboratory-derived embryonic and
induced-pluripotent stem cells) can give rise to all cell types of
the adult body. \engterm{Multipotent} cells (adult stem cells in the
bone marrow, intestinal crypt, hair follicle, etc.) can give rise to
a restricted set of related cell types. \engterm{Unipotent} cells
(spermatogonial stem cells, basal keratinocytes) can give rise to
only one differentiated cell type. The hierarchy maps onto the
tissue-engineering use case: pluripotent cells supply the broadest
range of cell types but require more careful control; multipotent
cells from the relevant tissue supply more lineage-appropriate cells
but require donor-tissue access.

\engterm{Induced pluripotent stem cells (iPSCs)}. Shinya Yamanaka's
2006 demonstration \cite{paper:v6c05-yamanaka-2006} that adult cells
could be reprogrammed to a pluripotent state by the expression of
four transcription factors (Oct3/4, Sox2, Klf4, c-Myc) transformed
the stem-cell field by removing the need for embryonic tissue. The
original mouse-fibroblast experiment used retroviral delivery of the
four factors and produced iPSC colonies at an efficiency of order
$10^{-3}$ to $10^{-5}$ of starting cells. Human iPSC derivation
followed in 2007. The Nobel Prize for the work was awarded in 2012.
iPSCs are now routinely derived from patient skin or blood cells and
differentiated into therapeutic cell types for autologous
transplantation. The four-factor system has been variously
re-engineered to eliminate integration risk: non-integrating
plasmid, Sendai-virus, and synthetic-mRNA reprogramming methods are
now the clinical standard (the full method matrix is in
\path{data/ipsc_reprogramming_efficiency.csv}). Clinical applications
using iPSC-derived cells include retinal pigment epithelium for
macular degeneration, cardiomyocytes for heart-failure patches
\cite{paper:v6c05-laflamme-cardiomyocyte-2007}, and dopaminergic
neurons for Parkinson's disease, with the first clinical trials
having begun in the mid-2010s and continuing at the time of writing.
Figure~\ref{fig:vol06ch05:yamanaka} shows the reprogramming and
differentiation paths.

\input{volumes/06-life/ch05-tissue-and-organ-engineering/figures/fig-yamanaka-reprogramming.tex}

\engterm{Worked example: the iPSC reprogramming cocktail}. The
working iPSC protocol takes a $4\,\text{mm}$ skin punch biopsy
(approximately $10^4$--$10^5$ primary dermal fibroblasts after a
two-week explant outgrowth) and reprograms it via Sendai-virus
delivery of OSKM (Oct3/4, Sox2, Klf4, c-Myc) over three to four
weeks. The Sendai vector is preferred for clinical work because it
does not integrate into the genome and clears spontaneously by
passage seven to ten. Reprogramming efficiency is $0.1\,\%$ to
$1\,\%$ of starting fibroblasts, giving $10^2$--$10^3$ iPSC colonies
per biopsy. Each colony is expanded over four to six additional
weeks to a clonal line of $10^7$--$10^8$ cells; the full sample-to-
clonal-line timeline is twelve to sixteen weeks. The clonal line is
then directed to the target differentiated lineage by a tissue-
specific protocol (cardiomyocyte differentiation requires Wnt
activation followed by Wnt inhibition over fourteen days; retinal
pigment epithelium requires Activin and Noggin over ninety days;
dopaminergic neurons require SHH and FGF8 over twenty-five days)
\cite{paper:v6c05-laflamme-cardiomyocyte-2007}. The fully-prepared
clinical-grade cell product takes approximately five to nine months
from patient biopsy to release, which is the working timeline for
the autologous iPSC clinical workflow.

\engterm{Organoids}. An \engterm{organoid} is a self-organised
three-dimensional culture of stem-cell-derived cells that
recapitulates aspects of an organ's structure and function.
Intestinal organoids (Hans Clevers's 2009 demonstration of
crypt-villus structures from single Lgr5-positive intestinal stem
cells \cite{paper:v6c05-clevers-organoids-2009}), brain organoids
(Madeline Lancaster's 2013 demonstration of cerebral organoids from
iPSCs \cite{paper:v6c05-lancaster-cerebral-2013}), kidney organoids,
liver organoids, and lung organoids are all now standard laboratory
models. Organoids do not reproduce full-organ function (they lack
vasculature, immune cells, and appropriate boundary tissues), but
they are valuable in disease modelling, drug screening, and as
building blocks for larger tissue constructs.

\engterm{The size limit of an avascular organoid}. The diffusion
limit derived above sets a ceiling on organoid growth. A cerebral
organoid grown in static culture reaches $2$ to $4\,\text{mm}$
diameter before the inner core necroses for want of oxygen; the
outer $\sim 200\,\si{\micro\meter}$ shell remains viable. Bioreactor-
spun cultures (the Lancaster 2013 protocol uses a spinning
bioreactor for medium exchange) extend the viable diameter to
$\sim 4\,\text{mm}$ but do not bypass the diffusion limit, only
reduce the boundary-layer resistance. Vascularising organoids is an
active area of research; the Takebe 2013 iPSC-derived liver bud
\cite{paper:v6c05-takebe-liver-2013} is the prototypical example of
co-culture-induced vascularisation. The reader should treat organoid
data on a tissue at $\le 4\,\text{mm}$ scale as informative for cell
phenotype but not for organ-level integration.

\engterm{Engineering control}. The challenge in stem-cell engineering
is controlling the differentiation: directing the cell to the desired
fate while suppressing alternative fates. The protocols are
combinations of soluble factors, ECM signals, mechanical cues, and
timing. Protocols are typically two to four weeks long and produce
populations of $\sim 80\,\%$ to $99\,\%$ purity of the target cell
type, with the remaining cells comprising off-target lineages or
undifferentiated cells. The undifferentiated cells in particular are
a concern for clinical use because of their teratoma-forming
potential; purification by surface-marker sorting before
transplantation is standard in current clinical protocols.

\begin{archetype}[Scaling and interface at the tissue scale]
Tissue engineering exhibits the scaling and interface archetypes in
their canonical clinical-translation forms. The scaling archetype is
the hierarchical assembly from molecular to cellular to tissue to
organ scales, with each transition requiring a new transport
mechanism (diffusion across the membrane, diffusion across the tissue,
flow through the vasculature). The interface archetype is the
necessity of designing every boundary in the construct: the
cell-matrix interface (integrin signalling, mechanotransduction), the
tissue-vasculature interface (endothelial lining, anti-thrombotic
function), the construct-host interface (immune compatibility,
vascular anastomosis, tissue integration). Tissue engineering is the
volume's most explicit application of the two archetypes to a single
engineering programme.
\end{archetype}

\section{Failure: rejection, vascularisation failure, tissue death}\pathtag{core}

Tissue-engineering constructs fail by three principal mechanism
classes, each well-characterised in translational experience.

\engterm{Immune rejection}. An implanted tissue construct that does
not match the recipient's HLA (human leukocyte antigen) type elicits
an immune response from the recipient's T-cells. The classical
classification, derived from solid-organ transplantation experience,
divides rejection into four time-course classes
(Figure~\ref{fig:vol06ch05:rejection}). \emph{Hyperacute rejection}
occurs within minutes to hours, mediated by pre-formed antibodies
(typically anti-ABO or anti-HLA from prior pregnancy, transfusion,
or transplant) that bind the donor endothelium and activate
complement and coagulation. Hyperacute rejection
eliminates the construct on perfusion and is the reason every
solid-organ transplant runs an ABO compatibility check and an
anti-HLA cross-match before implantation. \emph{Acute cellular
rejection} occurs over days to weeks, mediated by recipient T-cells
that recognise donor HLA molecules either directly (on donor cells)
or indirectly (on recipient antigen-presenting cells displaying
processed donor peptides). \emph{Acute antibody-mediated rejection}
operates over days to months and is driven by de novo
donor-specific antibodies that develop in the recipient. \emph{Chronic
rejection} is the slow vascular-wall fibrosis and luminal narrowing
that, despite immunosuppression, accumulates over months to years
and is the dominant cause of long-term graft loss. Strategies to
reduce rejection include HLA matching (the basis of solid-organ
transplant compatibility), pharmacological immunosuppression
(life-long for solid-organ recipients), autologous cell sourcing
(iPSC-derived cells from the patient), and immune-cloaked
constructs (genetically engineered cells with reduced HLA expression).
The immune-rejection failure mode constrains tissue engineering to a
working set of choices: autologous-source constructs avoid the
problem but require months of cell preparation per patient;
allogeneic constructs require either HLA matching or
immunosuppression. Chapter~\ref{vol06:ch12} on biocompatibility
develops the full biological-response engineering required for
medical devices, including tissue constructs.

\input{volumes/06-life/ch05-tissue-and-organ-engineering/figures/fig-rejection-pathways.tex}

\engterm{Vascularisation failure}. The implanted construct that fails
to vascularise within days of implantation suffers core necrosis.
The construct's inner cells, beyond the diffusion limit from the
surface, die from oxygen and nutrient starvation before the host
vasculature can grow into them. The dying cells release damage-
associated molecular patterns that recruit immune cells, accelerate
tissue breakdown, and may compromise the surviving outer shell. The
failure mode is the primary reason that solid-organ tissue
engineering has stalled at the centimetre scale for the past three
decades. The angiogenic ingrowth rate in vivo is approximately
$0.5\,\text{mm/day}$ for the wound-healing vasculature; a
$1\,\text{cm}$ thick construct must therefore survive on diffusion
alone for at least ten days while host vessels penetrate from the
surface, which is incompatible with the $\sim 200\,\si{\micro\meter}$
diffusion limit for the inner cells. The bypass strategies
(pre-vascularisation, channel-based design, sacrificial-template
fabrication) all aim at shortening this period by pre-installing
either microvessels or perfusable channels.

\engterm{Tissue death and remodelling failure}. Even a successfully
vascularised construct may fail to integrate into a working tissue
because the cells do not survive their first weeks in vivo, do not
remodel the scaffold into a native tissue, or do not maintain their
differentiated state. The mechanisms include inflammatory injury
from the foreign scaffold, mechanical stress that the construct is
not engineered to bear, and gradual loss of cellular phenotype
(dedifferentiation) as the cells adapt to the implant environment.
The three principal cell-death pathways operate on different
timescales and with different consequences: \emph{apoptosis} is the
programmed, energy-dependent, immunologically quiet death pathway
\cite{paper:v6c05-roberts-apoptosis-1972} in which the cell
fragments into membrane-bound apoptotic bodies and is cleared by
phagocytes without releasing immunogenic content; \emph{necrosis} is
the unprogrammed, energy-independent death pathway in which the
plasma membrane ruptures, releasing the cytoplasmic contents
(damage-associated molecular patterns: HMGB1, ATP, uric acid) that
recruit neutrophils and macrophages and amplify the inflammatory
response; \emph{autophagy} is the cell's self-cannibalisation
pathway that can either rescue the stressed cell or proceed to
death depending on the severity of stress. In tissue-engineering
failure, the dominant pathway depends on the mechanism: ischemic
core necrosis produces necrotic, immunogenic death; rejection-
mediated death is dominantly T-cell-mediated apoptosis early and
necrosis as the response amplifies; long-term cellular phenotypic
drift is a non-death failure mode that produces fibrosis rather
than tissue loss. Long-term clinical follow-up data on
tissue-engineered cartilage implants and bone-substitute scaffolds
show that a substantial fraction of constructs survive at $10$-year
follow-up but with remodelled rather than native tissue properties.

\engterm{The Macchiarini case as a worked failure analysis}. The
synthetic-trachea programme of 2011-2014
\cite{acc:macchiarini-trachea-2008-2014} illustrates all three
failure-mode classes operating simultaneously. The
$\sim 2\,\text{mm}$ thick polyethylene-terephthalate / polyurethane
scaffold walls were entirely above the diffusion limit; with no
pre-vascularisation and only forty-eight hours of bone-marrow-cell
seeding before implantation, the inner cellular content of the
scaffold either died or never became established. Post-mortem
histology confirmed that the scaffold walls remained acellular
through their bulk \cite{paper:v6c05-karolinska-heckscher-2018,
paper:v6c05-delaere-2014-thorax}. Vascularisation failure was
therefore the dominant mechanism. The synthetic polymer also failed
the host-integration interface: the recipient's connective tissue
did not bond to the polymer surface, producing a separation at the
scaffold-host interface that progressed to mediastinal fistula and
vascular erosion. The cellular-death pathway was dominantly necrosis
of the seeded cells plus a chronic inflammatory response to the
synthetic polymer surface, producing granulation tissue and
mediastinitis. The clinical-outcome consequence was uniformly fatal
in seven of eight cases. The engineering-chain consequence:
the absence of preclinical animal-model validation meant that the
failure mechanisms predicted by Krogh-cylinder diffusion limits, by
the Engler stiffness-matched-substrate principle, and by the basic
vascularisation arithmetic above were not tested before patient
implantation. The case is the worked example of what the
diffusion-limit, stiffness-matching, and host-integration
principles of this chapter say when ignored.

\engterm{Engineering responses}. Each failure mode has well-characterised
engineering responses: HLA matching or autologous sourcing for
rejection; pre-vascularisation or microchannel design for
vascularisation; scaffold stiffness matching and bioactive coating
for tissue integration. The engineering of tissue constructs is, in
practice, an ongoing process of preempting these three failure modes
across the construct's geometry, materials, cell source, and
implantation protocol. The Folkman 1971 angiogenic-recruitment
hypothesis \cite{paper:v6c05-folkman-1973-tumour-angiogenesis}
remains the theoretical underpinning of the host-vessel-ingrowth
strategy; the Murray and Krogh laws set the geometric and transport
constraints; Engler's matrix-stiffness principle sets the
cell-fate constraint. A construct designed against all three is a
candidate for translation; a construct that ignores any one of them
fails predictably.

\section{Project}\pathtag{core}

\begin{project}[Design a vascularised scaffold and defend the geometry]
The reader designs a vascularised scaffold for a specified
tissue-engineering target and produces a written defence of the
chosen geometry, materials, and vascularisation strategy.

\textbf{Track.} Design study (paper exercise, no fabrication required).
\textbf{Hazard class}: none.

\textbf{Target.} The reader selects one of three target tissues:
(a) a $2\,\text{cm} \times 2\,\text{cm} \times 0.5\,\text{cm}$ skin
substitute for a chronic non-healing wound; (b) a
$4\,\text{cm} \times 4\,\text{cm} \times 1\,\text{cm}$ cartilage
construct for a knee joint repair; (c) a
$5\,\text{cm} \times 5\,\text{cm} \times 2\,\text{cm}$ liver-tissue
patch for partial hepatic regeneration after surgical resection.

\textbf{Procedure.}

\begin{enumerate}[leftmargin=*]
\item Specify the target tissue's mechanical properties (Young's
  modulus, strain to failure), cellular composition, and vasculature
  requirement.
\item Choose a scaffold material and defend the choice on the basis
  of degradation rate, mechanical match, and cell-attachment
  signals.
\item Design the macroscopic geometry, including any internal
  channels for vascularisation. Specify channel diameters, spacing,
  and the strategy for connecting to host circulation.
\item Specify the cell source (autologous, allogeneic, iPSC-derived)
  and the seeding density.
\item Identify the three most likely failure modes and the engineering
  responses you have built in.
\item Estimate the cost per construct (rough order-of-magnitude) and
  the time from patient sample to implantation.
\end{enumerate}

\textbf{Deliverable.} A ten- to fifteen-page report with one geometric
diagram, one materials table, one failure-mode analysis table, and a
closing paragraph naming the one engineering uncertainty most likely
to limit the design.

\textbf{Time.} Eight to twelve hours.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Compute the maximum thickness of a metabolically active tissue
($\dot{q} = 10^{-3}\,\text{mol O}_2/(\text{m}^3\cdot\text{s})$)
supplied by diffusion from one face, with $D_{\text{O}_2} = 2 \times
10^{-9}\,\text{m}^2/\text{s}$ and $[\text{O}_2]_{\text{surface}} =
0.13\,\text{mM}$.
\end{exercise}

\paragraph{Solution sketch.} The steady-state oxygen-balance equation
through a slab of thickness $L$, with one-face supply at
concentration $C_0$, and a far-face condition that the cell is at
the threshold of starvation ($C(L) = 0$, $\dd C/\dd x|_{L} = 0$
simultaneously), has the closed-form solution
\[
L_{\max} = \sqrt{\frac{2 D C_0}{\dot{q}}}.
\]
Substituting $D = 2 \times 10^{-9}\,\text{m}^2/\text{s}$,
$C_0 = 0.13\,\text{mM} = 0.13\,\text{mol/m}^3$, and
$\dot{q} = 10^{-3}\,\text{mol/(m}^3\,\text{s)}$ gives
\[
L_{\max} = \sqrt{\frac{2 \times 2 \times 10^{-9} \times 0.13}{10^{-3}}}\,\text{m}
        = \sqrt{5.2 \times 10^{-7}}\,\text{m}
        \approx 230\,\si{\micro\meter}.
\]
The number is within a factor of two of the working
$\sim 200\,\si{\micro\meter}$ figure that the chapter cites; it is the
slab-geometry upper bound. The Krogh-cylinder geometry gives a
slightly smaller number ($\sim 100$ to $200\,\si{\micro\meter}$
inter-capillary spacing) because the capillary supplies a surrounding
cylinder of tissue rather than a half-space.

\begin{exercise}
Murray's law predicts that $d_{\text{parent}}^3 = d_1^3 + d_2^3$ at
a vascular bifurcation. For a symmetric bifurcation where
$d_1 = d_2$, compute the ratio $d_{\text{daughter}}/d_{\text{parent}}$.
\end{exercise}

\paragraph{Solution sketch.} Setting $d_1 = d_2 = d_d$ in Murray's
law gives
\[
d_p^3 = 2 d_d^3, \quad
\frac{d_d}{d_p} = 2^{-1/3} \approx 0.7937.
\]
For a tree of $n$ symmetric bifurcations starting from a root of
diameter $d_0$, the leaf diameter is $d_n = d_0 \cdot 2^{-n/3}$.
Reducing an aortic root of $25\,\text{mm}$ to a capillary of
$8\,\si{\micro\meter}$ requires
$n = 3 \log_2(25\,\text{mm}/8\,\si{\micro\meter}) \approx 35$
symmetric bifurcations (lower bound; real vascular trees mix
symmetric and asymmetric branching). The ratio $2^{-1/3}$ is also the
working diameter rule in the bioprinting literature for
sacrificial-template channel networks.

\begin{exercise}
A $1\,\text{cm}^3$ tissue construct has a vasculature volume
fraction of $10\,\%$ and a capillary density of $500\,\text{mm}^{-2}$.
Compute the total capillary length in the construct, assuming
$8\,\si{\micro\meter}$ capillary diameter.
\end{exercise}

\paragraph{Solution sketch.} The vasculature volume is
$V_{\text{vasc}} = 0.10 \times 1\,\text{cm}^3 = 0.1\,\text{cm}^3
= 10^{-7}\,\text{m}^3$. Treat all of it as capillaries (a slight
overestimate; the larger vessels contribute $\sim 5\,\%$). Capillary
cross-sectional area at $8\,\si{\micro\meter}$ diameter is
$A_c = \pi (4 \times 10^{-6})^2 \approx 5.03 \times 10^{-11}\,\text{m}^2$.
The total capillary length is
\[
L = \frac{V_{\text{vasc}}}{A_c}
  = \frac{10^{-7}}{5.03 \times 10^{-11}}
  \approx 2000\,\text{m} = 2\,\text{km}.
\]
The cross-check using the stated capillary density:
$500\,\text{mm}^{-2}$ over the $1\,\text{cm}^2$ cross-section of a
$1\,\text{cm}^3$ cube is $5 \times 10^4$ capillaries, each running
the $1\,\text{cm}$ depth, giving $500\,\text{m}$ of capillary per
slice. A $1\,\text{cm}^3$ cube is one slice, so the geometric
density figure undercounts by a factor of $4$ because real
capillaries branch and lattice. Both numbers are working
order-of-magnitude estimates; the right answer in a real construct
is a histology measurement, not a closed-form derivation.

\begin{exercise}
Tendon has a Young's modulus of $\sim 1\,\text{GPa}$ along the
fibre and breaks at a tensile stress of $\sim 100\,\text{MPa}$.
Compute the strain at failure and compare to skin (modulus $\sim 0.1\,\text{MPa}$
at small strain, failure at strain $\sim 0.5$).
\end{exercise}

\paragraph{Solution sketch.} For tendon in the linear regime, strain
at failure is $\varepsilon_{\text{fail}} = \sigma_{\text{fail}}/E
= 10^{8}\,\text{Pa} / 10^{9}\,\text{Pa} = 0.1$, so tendon breaks at
approximately $10\,\%$ strain. The small-strain modulus is a poor
estimator for the failure strain of strain-stiffening materials;
in tendon the failure regime is approximately linear because the
toe region is narrow. Skin, by contrast, has a small-strain modulus
of $\sim 0.1\,\text{MPa}$ but exhibits a sharp J-curve such that the
secant modulus at $50\,\%$ strain is two orders of magnitude
larger; computing strain at failure from the small-strain modulus
would predict $\varepsilon = 10^{8}/10^{5} = 1000$, an unphysical
answer that highlights why constitutive modelling matters. Skin's
true failure strain is $\sim 0.5$ (a factor of five larger than
tendon's), but skin's failure stress is also $\sim 15\,\text{MPa}$
(a factor of seven smaller than tendon's), so skin and tendon
distribute their tensile load in qualitatively different ways:
tendon is stiff and high-strength with little extensibility, skin
is compliant and lower-strength with substantial extensibility.

\begin{exercise}
A PLGA scaffold has an in vivo degradation half-life of $4\,\text{weeks}$.
Compute the mass fraction remaining at $1\,\text{week}$, $4\,\text{weeks}$,
and $12\,\text{weeks}$ assuming first-order degradation kinetics.
\end{exercise}

\paragraph{Solution sketch.} First-order degradation gives
$m(t)/m_0 = (1/2)^{t/t_{1/2}} = e^{-(\ln 2 / t_{1/2}) t}$.
With $t_{1/2} = 4\,\text{weeks}$:
\[
\begin{aligned}
\frac{m(1\,\text{wk})}{m_0} &= 2^{-1/4} \approx 0.84,\\
\frac{m(4\,\text{wk})}{m_0} &= 2^{-1} = 0.50,\\
\frac{m(12\,\text{wk})}{m_0} &= 2^{-3} = 0.125.
\end{aligned}
\]
The engineering implication: at three months in vivo the
PLGA-50:50 scaffold has degraded to one-eighth of its starting
mass, with the corresponding loss of mechanical strength (strength
scales faster than mass because surface erosion concentrates load
on the remaining structure). The scaffold's
mechanical-strength-vs-time profile must therefore match the host's
tissue regeneration timeline: too-fast degradation produces
mechanical failure before host tissue is established; too-slow
degradation leaves a foreign body in place after regeneration is
complete. PLGA blends with different lactic / glycolic ratios
($50{:}50$ for fast, $85{:}15$ for slow) are how the designer
tunes the time constant.

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the diffusion-limited maximum tissue thickness from the
steady-state oxygen-balance equation
$D \nabla^2 C = \dot{q}$ with boundary conditions
$C(0) = C_{\text{surface}}$, $C(L) = 0$ for a slab of thickness $L$
supplied from one face.
\end{exercise}

\paragraph{Solution sketch.} In one dimension the steady-state
equation is
\[
D \frac{\dd^2 C}{\dd x^2} = \dot{q}.
\]
Integrate twice:
\[
\frac{\dd C}{\dd x} = \frac{\dot{q}}{D} x + A, \quad
C(x) = \frac{\dot{q}}{2D} x^2 + A x + B.
\]
Apply boundary conditions $C(0) = C_0$ (giving $B = C_0$) and
$C(L) = 0$:
\[
0 = \frac{\dot{q}}{2D} L^2 + A L + C_0
\Rightarrow A = -\frac{\dot{q} L}{2 D} - \frac{C_0}{L}.
\]
The maximum sustainable $L$ is the thickness at which the
concentration minimum (in the slab interior) is exactly zero; for
the supply-from-one-face case the minimum sits at the far face when
$\dd C/\dd x |_{L} = 0$ simultaneously, giving the additional
condition
\[
\frac{\dot{q}}{D} L + A = 0 \Rightarrow A = -\frac{\dot{q} L}{D}.
\]
Combining the two expressions for $A$:
\[
-\frac{\dot{q} L}{D} = -\frac{\dot{q} L}{2 D} - \frac{C_0}{L},
\]
which solves to $L_{\max}^2 = 2 D C_0 / \dot{q}$, i.e.\
$L_{\max} = \sqrt{2 D C_0 / \dot{q}}$. For oxygen in metabolically
active tissue this is $\approx 230\,\si{\micro\meter}$. The
derivation generalises: any first-order steady-state diffusion-
consumption problem produces the same $\sqrt{D C_0 / \dot{q}}$
structure, with a geometry-dependent prefactor.

\begin{exercise}
Derive Murray's law for the optimum diameter ratio at a vascular
bifurcation by minimising the total cost (metabolic cost of
maintaining the vessel, scaling with volume, plus the work cost of
flow, scaling with the inverse of $d^4$).
\end{exercise}

\paragraph{Solution sketch.} Consider a vessel of length $\ell$ and
diameter $d$ carrying flow rate $Q$ (volume per unit time). The
total cost has two components.

\emph{Maintenance cost.} Blood is metabolically expensive to
maintain; its cost scales with vessel volume,
$C_m = k_m \cdot \ell \cdot d^2$, with $k_m$ the cost per unit
volume per unit time.

\emph{Pumping cost.} Steady laminar flow through a cylindrical pipe
follows Poiseuille's law $\Delta P = 128 \mu \ell Q / (\pi d^4)$.
The pumping power is $P_{\text{pump}} = Q \cdot \Delta P =
128 \mu \ell Q^2 / (\pi d^4)$. Identify the pumping cost as
$C_p = k_p \cdot \ell \cdot Q^2 / d^4$ with $k_p = 128 \mu / \pi$.

The total cost is $C = k_m \ell d^2 + k_p \ell Q^2 / d^4$.
Minimising with respect to $d$:
\[
\frac{\dd C}{\dd d} = 2 k_m \ell d - 4 k_p \ell Q^2 / d^5 = 0
\Rightarrow d^6 = \frac{2 k_p Q^2}{k_m}
\Rightarrow d \propto Q^{1/3}.
\]
At a bifurcation, $Q_{\text{parent}} = Q_1 + Q_2$, so
$d_{\text{parent}}^3 = d_1^3 + d_2^3$, which is Murray's law. The
symmetric case $Q_1 = Q_2$ gives
$d_{\text{daughter}} = 2^{-1/3} d_{\text{parent}}$. The same
optimisation, extended to the angle, gives the
$\theta \approx 37.5^{\circ}$ result for the symmetric optimum.

\begin{exercise}
Derive the strain-energy density function for a Mooney-Rivlin
material in terms of the strain invariants $I_1$ and $I_2$, and
identify the uniaxial-tension stress response.
\end{exercise}

\paragraph{Solution sketch.} The Mooney-Rivlin strain-energy
function for an incompressible material is
\[
W(I_1, I_2) = c_1 (I_1 - 3) + c_2 (I_2 - 3),
\]
with $I_1 = \lambda_1^2 + \lambda_2^2 + \lambda_3^2$ and
$I_2 = \lambda_1^2\lambda_2^2 + \lambda_2^2\lambda_3^2
       + \lambda_1^2\lambda_3^2$ the first two strain invariants in
terms of principal stretches $\lambda_i$. The $-3$ shifts ensure
$W = 0$ at the reference state ($\lambda_i = 1$). For
incompressible uniaxial tension along axis 1, $\lambda_1 = \lambda$
and $\lambda_2 = \lambda_3 = 1/\sqrt{\lambda}$. Substituting:
$I_1 = \lambda^2 + 2/\lambda$,
$I_2 = 2\lambda + 1/\lambda^2$.

For an incompressible hyperelastic material in uniaxial tension the
Cauchy stress is
$\sigma = \lambda \,\partial W / \partial \lambda
       - p$
where the hydrostatic pressure $p$ enforces transverse stress-free
boundaries. Working through the algebra:
\[
\sigma(\lambda) = 2\!\left(c_1 + \frac{c_2}{\lambda}\right)
                  \!\left(\lambda^2 - \frac{1}{\lambda}\right).
\]
At $\lambda \to 1$ this gives the small-strain modulus $E_0 =
6(c_1 + c_2)$, fixing one combination of the parameters from the
small-strain measurement. Mooney-Rivlin requires two-point fitting
(e.g., small-strain modulus plus secant modulus at a specified
larger strain) to identify $c_1$ and $c_2$ separately. The model
captures stiffening up to a factor of $\sim 5$ before the
exponential Fung form becomes necessary.

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the number of cells required to populate a
$5\,\text{cm}^3$ liver tissue construct at native liver cellular
density ($\sim 10^8\,\text{cells}/\text{cm}^3$).
\end{exercise}

\paragraph{Solution sketch.} Direct multiplication:
$N = 10^8 \times 5 = 5 \times 10^8$ cells. Adjusting for the
vasculature reservation ($\sim 10\,\%$) and the non-parenchymal
cell complement (hepatocytes are $\sim 60\,\%$ by cell count, with
endothelial cells, stellate cells, and Kupffer cells making up the
balance), the hepatocyte requirement is $\sim 3 \times 10^8$ and
the non-parenchymal supply is $\sim 2 \times 10^8$. Sourcing the
hepatocytes is the binding constraint: primary human hepatocytes
have a doubling time of weeks and limited expansion capacity, so
the cells must come from either a large primary donor (fresh
hepatocytes from a marginal donor liver) or iPSC-derived
hepatocyte-like cells, which take twelve to sixteen weeks from
biopsy to clinical product. The cell-seeding-density script at
\path{code/cell_seeding_density.py} gives the full scale-up
arithmetic across construct sizes.

\begin{exercise}
Estimate the time required for vasculature to grow into an implanted
tissue construct of $1\,\text{cm}^3$, given a typical neoangiogenic
growth rate of $\sim 0.5\,\text{mm}/\text{day}$.
\end{exercise}

\paragraph{Solution sketch.} A $1\,\text{cm}^3$ cube has a centre
$5\,\text{mm}$ from the nearest surface; angiogenic vessels growing
from the surface inward at $0.5\,\text{mm/day}$ reach the centre
in $\sim 10\,\text{days}$. For the centre cells to survive that
ten-day interval on diffusion alone, the cells at depth
$5\,\text{mm}$ must remain viable for the period, which is
incompatible with the $\sim 200\,\si{\micro\meter}$ diffusion
limit. The arithmetic is the principal reason that
$\ge 1\,\text{cm}$ thick constructs without internal vasculature
(pre-vascularisation or perfusable channels) cannot survive
implantation. The bypass strategies aim to deliver host-vessel
perfusion within hours, not days, of implantation. The Miller 2012
sacrificial-template approach \cite{paper:v6c05-miller-vascular-print-2012}
produces a perfusable channel network from day zero; the Takebe
2013 liver bud \cite{paper:v6c05-takebe-liver-2013} relies on
co-culture-induced pre-vascularisation; both compress the
host-vessel-ingrowth period to hours.

\begin{exercise}
Estimate the cost (in dollars) of producing an iPSC-derived
cardiomyocyte population of $10^9$ cells, given current commercial
prices for the reagents and growth media (current as of 2025).
\end{exercise}

\paragraph{Solution sketch.} The dominant cost components are
growth media, growth factors, and the bioreactor consumables.
For an iPSC-to-cardiomyocyte differentiation: the iPSC expansion
phase (from $10^6$ to $10^9$ cells, ten doublings) takes ten days
in mTeSR1 or equivalent xeno-free medium at $\sim \$200/\text{L}$,
consuming $\sim 50\,\text{L}$ of medium over the expansion
($\$10{,}000$). The differentiation phase (cardiomyocyte protocol)
takes fourteen to twenty days and consumes growth factors
(Activin~A at $\sim \$3{,}500/\text{mg}$, BMP4 at
$\sim \$1{,}500/\text{mg}$, CHIR99021 at $\sim \$500/\text{g}$),
giving $\sim \$15{,}000$ in growth factors plus another
$\$15{,}000$ in differentiation medium. Quality-control assays
(flow cytometry, contractility imaging, electrophysiology) add
$\sim \$5{,}000$. The total is $\$45{,}000$ to $\$60{,}000$ per
$10^9$-cell batch under research-grade conditions; under
GMP-grade conditions the cost rises to $\$150{,}000$ to
$\$300{,}000$ per batch because the medium grade, the cleanroom
overhead, the QC battery, and the documentation are
multiplicatively more expensive. The unit cost of GMP-grade
iPSC-derived cardiomyocytes at clinical scale is therefore
$\sim \$0.15$ to $\$0.30$ per cell, and a $10\,\text{g}$ cardiac
patch ($10^9$ cells) is a $\$150{,}000$ to $\$300{,}000$
manufacturing item before any clinical, infrastructure, or
regulatory overhead. The number is the binding constraint on the
near-term clinical economics of iPSC therapy and is the principal
reason that early indications are for high-value, low-cell-count
applications (retinal pigment epithelium for macular degeneration:
$\sim 10^5$ cells per dose) rather than high-cell-count
applications (cardiac infarct repair: $\sim 10^9$ cells per dose).

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Simulate the diffusion of oxygen through a slab of metabolically
active tissue using a finite-difference solver. Verify the analytic
maximum-thickness limit and identify the sensitivity to the diffusion
coefficient and the consumption rate.
\end{exercise}

\paragraph{Solution sketch.} A working implementation is provided
at \path{code/krogh_oxygen_diffusion.py}. The reader sets up a
uniform finite-difference grid on $[0, L]$ with Dirichlet boundary
$C(0) = C_0$ and Dirichlet boundary $C(L) = 0$, discretises the
balance equation as
$D (C_{i+1} - 2 C_i + C_{i-1})/\Delta x^2 = \dot{q}$, solves the
tridiagonal system, and confirms that the minimum interior
concentration is zero when $L = \sqrt{2 D C_0 / \dot{q}}$. The
sensitivity sweep over $D \in [5 \times 10^{-10}, 10^{-8}]\,\text{m}^2/\text{s}$
and $\dot{q} \in [10^{-4}, 10^{-2}]\,\text{mol/(m}^3\,\text{s)}$
shows $L_{\max}$ scaling as $\sqrt{D/\dot{q}}$, a factor of $30$
range across the physiological span: from a $\sim 60\,\si{\micro\meter}$
limit in highly active cardiac muscle to a $\sim 2\,\text{mm}$
limit in low-consumption avascular cartilage. The rubric for the
exercise: a passing solution reports the analytic-numerical
agreement at $L = L_{\max}$, the qualitative sweep behaviour, and
the implications for cartilage (which is uniquely diffusion-supplied
in vivo because of its low cellular density and low metabolic
rate).

\begin{exercise}
Implement a finite-element model of a Mooney-Rivlin material in
uniaxial tension. Fit the model to experimental stress-strain data
for a soft tissue and identify the limits of the fit.
\end{exercise}

\paragraph{Solution sketch.} A reduced one-element scalar
implementation suffices for uniaxial tension and is provided at
\path{code/tissue_stress_strain_fit.py}. The script generates
synthetic Fung-like stress-strain data and fits neo-Hookean,
Mooney-Rivlin, and Fung forms by nonlinear least-squares (SciPy
\verb|curve_fit|). The expected outcome: neo-Hookean fits poorly
(RMSE $\sim 10\times$ the Fung RMSE) because it has no
strain-stiffening capacity; Mooney-Rivlin fits moderately well
(RMSE $\sim 3\times$ the Fung RMSE) because the $c_2/\lambda$
term gives a modest stiffening; Fung fits to within noise because
the data is Fung-generated. On real tissue data, the qualitative
ordering holds: neo-Hookean is inadequate for any tissue with a
J-curve; Mooney-Rivlin is adequate up to a stiffening factor of
$\sim 5$; Fung covers the full physiological range of skin, artery,
and tendon. The rubric: a passing submission reports the three RMSE
values, the fitted parameters, and the diagnostic that Mooney-Rivlin
fails at a stiffening factor above approximately $5$. Extension
exercises: implement the HGO model
\cite{paper:v6c05-holzapfel-hgo-2000} for biaxial arterial-wall
data and report the fibre-orientation parameter.

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a porous scaffold for bone regeneration. Specify the pore
size, the wall stiffness, and the material. Defend the choice
against the alternative options (smaller pores, stiffer walls,
different materials).
\end{exercise}

\paragraph{Discussion.} The working specification for a bone-
regeneration scaffold: pore size $200$ to $400\,\si{\micro\meter}$
(the bone-ingrowth window; pores $\le 100\,\si{\micro\meter}$ slow
vascular ingrowth and pores $\ge 500\,\si{\micro\meter}$ reduce
osteoblast attachment density); modulus matched to the host bone
cortex ($\sim 15\,\text{GPa}$); porosity $60$ to $80\,\%$ (the
trade-off window between mechanical adequacy and pore
connectivity); interconnectivity throat $\ge 50\,\si{\micro\meter}$
to support vascular sprouts; material a porous hydroxyapatite or
PLGA-ceramic composite that mineralises into bone over months.
Defence: smaller pores ($50$--$100\,\si{\micro\meter}$) would
restrict cell infiltration and slow vascularisation, leading to
core necrosis; stiffer walls (a fully dense ceramic) would
violate the Wolff-law-mediated remodelling logic (the host bone
adapts its density to the load it carries; an inelastic implant
stress-shields the surrounding bone and leads to peri-implant
osteopenia); a pure PLA / PLGA scaffold without a ceramic
component would lack the osteoinductive surface chemistry that
recruits osteoblast attachment. A passing submission references
the pore-size design window, the modulus-matching argument, and
the stress-shielding mechanism, and proposes a measurement to
verify that the host bone is loading the scaffold (e.g., DEXA at
six-month follow-up).

\begin{exercise}
Design a hydrogel-based scaffold for neural tissue engineering.
Specify the polymer chemistry, the stiffness (target $\sim 1\,\text{kPa}$),
and the cell-attachment ligands.
\end{exercise}

\paragraph{Discussion.} The working specification: a methacrylated-
gelatin (GelMA) hydrogel at $5\,\%$ to $10\,\%$ w/v with a low
photoinitiator concentration, cross-linked to a target modulus of
$1$ to $3\,\text{kPa}$ to match brain-tissue stiffness and bias
mesenchymal-stem-cell differentiation toward neurons (the Engler
2006 principle); cell-attachment ligands RGD (fibronectin-derived)
plus IKVAV or YIGSR (laminin-derived) at micromolar surface density
to support neurite extension; pore size $20$ to $50\,\si{\micro\meter}$
to allow axon and dendrite extension without imposing a rigid
template. Defence: a stiffer scaffold ($> 5\,\text{kPa}$) would
bias mesenchymal cells toward muscle or osteoblast lineages, away
from the desired neuronal fate; a pure polyethylene-glycol hydrogel
without peptide ligands would fail to support neural-cell
attachment; alginate would lack the protease-cleavable backbone
needed for remodelling-mediated integration. The submission should
quantify the modulus measurement (rheometer or AFM nanoindentation)
and the ligand density (ELISA or surface-plasmon-resonance).

\begin{exercise}
Design a vascular graft for replacing a $10\,\text{cm}$ length of
femoral artery. Specify the diameter, wall thickness, mechanical
properties, and the strategy for managing thrombosis on the
luminal surface.
\end{exercise}

\paragraph{Discussion.} The working specification: outer diameter
$6\,\text{mm}$ (matching the host femoral artery), wall thickness
$0.5\,\text{mm}$, compliance matched to native artery
(circumferential modulus $\sim 0.5\,\text{MPa}$ at low strain,
strain-stiffening factor of $\sim 5$ at $\varepsilon = 0.2$),
length $10\,\text{cm}$ plus a $1\,\text{cm}$ allowance at each
anastomosis. Wall structure: an outer electrospun PCL layer for
mechanical strength, a middle decellularised arterial ECM layer
for vascular-smooth-muscle attachment, an inner endothelial-
cell-seeded layer to manage thrombosis. Anti-thrombotic strategy:
the endothelial monolayer is the primary defence (intact
endothelium suppresses platelet activation and clotting); heparin
covalent attachment to the inner surface is a secondary defence
during the period before endothelialisation. Defence against
alternatives: a synthetic-polymer-only graft (ePTFE or Dacron, the
clinical standard for large-vessel replacement) is inadequate for
$\le 6\,\text{mm}$ vessels because synthetic-surface thrombosis
overwhelms the design; an autologous saphenous-vein graft is the
current clinical standard but requires harvesting a healthy donor
vein, with morbidity at the donor site. The proposed tissue-
engineered graft addresses both limitations but requires
demonstrating that the endothelial-cell monolayer survives the
arterial wall shear stress ($\sim 1$ to $2\,\text{Pa}$). A passing
submission specifies a shear-flow validation in a bioreactor and
the in-vivo monitoring strategy (Doppler ultrasound for patency at
six-month intervals).

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A tissue-engineered cartilage implant fails six months after
implantation; histology shows mostly fibrous tissue rather than
hyaline cartilage. Identify three possible engineering failure modes
(wrong cell source, wrong mechanical environment, insufficient
chondrogenic signalling) and propose a measurement to distinguish
them.
\end{exercise}

\paragraph{Discussion.} The three failure modes are dedifferentiation
of expanded autologous chondrocytes (cells switch from hyaline
chondrocyte phenotype to fibroblast-like phenotype during the in-vitro
expansion phase), inadequate compressive loading post-implantation
(hyaline cartilage requires daily cyclic compressive loading to
maintain phenotype; an unloaded or over-loaded scaffold drifts toward
fibrocartilage), and inadequate growth-factor support (TGF-$\beta$
and BMP-6 are required for chondrogenic maintenance, and a scaffold
that does not retain or release these factors fails to maintain the
phenotype). Discriminating measurements: (i)~retain a cryopreserved
aliquot of the cell preparation that was implanted, immunostain for
chondrogenic markers (Sox9, type II collagen) and fibroblast markers
(type I collagen), compare to the explant biopsy; if the implanted
cells were already fibroblast-like, dedifferentiation in vitro is
the cause. (ii)~Quantify the patient's joint-loading history
(accelerometer at home, gait analysis at clinic) and compare to the
intended rehabilitation protocol; under- or over-loading suggests
mechanical-environment failure. (iii)~Measure the local concentration
of TGF-$\beta$ at the explant site and compare to the controlled-
release predictions; below-target concentration suggests signalling
failure. A passing submission identifies the three failure modes,
proposes a distinguishing measurement for each, and notes the order
in which to run them (cheapest first: the cryopreserved-cell
phenotype check is free and excludes the most common cause).

\begin{exercise}
A pre-vascularised skin substitute integrates well at the edges but
shows central necrosis. Identify the engineering cause and propose
one design modification that would address it.
\end{exercise}

\paragraph{Discussion.} The edge-integration / central-necrosis
pattern is the signature of inadequate central vascularisation: at
the edges, host vessels grow in from the surrounding healthy tissue
within days (within the $0.5\,\text{mm/day}$ angiogenic-growth
budget for $\sim 2$ to $5\,\text{mm}$ edge zones), but the central
region requires either ingrowth from a longer distance (beyond the
ten-day diffusion-survival budget) or anastomosis between the
pre-vascular network and a host vessel. The pre-vascular network
formed in vitro evidently did not anastomose. Design modification:
add an external trunk-vessel inlet and outlet to the pre-vascular
network at the time of implantation, surgically anastomosed to a
host vessel (the strategy used for free-flap reconstruction). The
modification requires that the construct be designed with a
millimetre-scale trunk vessel for surgical anastomosis, beyond the
microcapillary network. Alternatively, the construct's
pre-vascular network can be ``primed'' for anastomosis by surface
expression of pro-anastomotic signals (VEGF, angiopoietin-1) at
the construct boundary, so that host vessels are recruited to the
edge of the pre-network and connect to it within days. A passing
submission identifies the anastomosis failure as the cause and
proposes a specific design modification (with the trade-off:
surgical-anastomosis adds an hour of operative time but provides
immediate perfusion).

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A patient receives an allogeneic tissue-engineered tracheal graft.
The graft is acutely rejected at three weeks despite immunosuppression.
Identify the most likely engineering cause and the chain of events
from immune recognition to graft destruction.
\end{exercise}

\paragraph{Discussion.} A three-week timeline points to acute
cellular rejection mediated by T-cells. The most likely engineering
cause is incomplete decellularisation of the allogeneic donor
trachea: residual donor cells expressing class-I and class-II HLA
molecules are recognised by recipient CD8+ and CD4+ T-cells, which
infiltrate the construct and lyse the donor cells. The chain of
events: (1)~donor HLA presented on residual donor cells; (2)~recipient
antigen-presenting cells (dendritic cells, macrophages) infiltrate
the construct and process donor antigens for cross-presentation;
(3)~recipient T-cells in regional lymph nodes are primed to donor
HLA; (4)~primed T-cells traffic to the graft, infiltrate the
construct, and lyse donor cells; (5)~the lysed donor cells release
damage-associated molecular patterns that amplify the inflammatory
response; (6)~the graft loses cellular content and structural
integrity. Diagnostic check: histology of the explanted graft for
T-cell infiltrate (CD8+, CD4+) and residual donor HLA expression;
flow cytometry of recipient blood for donor-specific antibodies;
quantification of immunosuppression drug levels to confirm the
prescribed dose was achieved. The engineering response: improve the
decellularisation protocol (SDS + Triton X-100 sequential
treatments; verify residual DNA $< 50\,\text{ng/mg}$ tissue by
PicoGreen assay) before patient use. The exercise is a worked
example of one of the three Macchiarini-case failure mechanisms,
inverted to its diagnostic form. A passing submission identifies
the residual donor antigen as the cause and the decellularisation-
quality measurement as the engineering response.

\begin{exercise}
A 3D-printed tissue construct fails mechanically (tears under
physiological strain) within $48\,\text{hours}$ of implantation.
Identify three possible engineering causes (wrong polymer choice,
wrong crosslinking density, premature degradation in vivo, inadequate
mechanical match to surrounding tissue) and propose how to
distinguish them.
\end{exercise}

\paragraph{Discussion.} Three candidate causes: (1)~insufficient
crosslinking density at the time of implantation (the photo-
crosslinked hydrogel was under-cured and the as-implanted modulus
was below the design specification); (2)~accelerated in-vivo
degradation (the polymer's in-vitro degradation rate was measured
under different pH or enzyme conditions than the implant
environment); (3)~mechanical mismatch with the surrounding tissue
(the construct is too stiff or too compliant for the surrounding
matrix, concentrating stress at the implant interface). Distinguishing
measurements: (i)~retain a slice of the original construct, run
nanoindentation and compare to the design specification; if the
as-printed modulus is already below target, under-curing is the
cause. (ii)~Explant a fragment, measure mass loss against the
design degradation curve; accelerated mass loss indicates
unanticipated enzymatic or chemical attack. (iii)~Image the
implant boundary at four hours and twenty-four hours; visible
stress concentrations at the interface (cracks initiating at the
boundary) indicate a mechanical mismatch failure. The order:
under-curing is the cheapest to test (nanoindentation on a retained
sample) and the most common cause in development-stage workflow.
A passing submission identifies the three failure modes,
distinguishes them with appropriate measurements, and discusses
the order in which to run the diagnostics.

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A press release announces a ``3D-printed kidney that filters blood
in vitro''. Identify the engineering claims (vascular completeness,
nephron function, immune compatibility) and write a one-paragraph
reply identifying the likely state of each at the time of the
press release.
\end{exercise}

\paragraph{Discussion.} A model reply: ``The press-release language
``3D-printed kidney that filters blood in vitro'' is sufficiently
ambiguous to cover a wide range of underlying technical claims. We
ask the team to specify three figures: (1)~the construct's volume
and the cell density it is loaded at, against the native kidney
target ($\sim 150\,\text{cm}^3$ at $\sim 5 \times 10^7$
nephron-cell-equivalents per $\text{cm}^3$, roughly $10^{10}$
nephron cells; we expect the construct is at $\sim 1\,\text{cm}^3$
and a small fraction of native cell density); (2)~the vascular
network completeness as measured by the perfusion fraction at
physiological pressure ($100\,\text{mmHg}$) and the diffusion
distance from the nearest perfused channel ($> 200\,\si{\micro\meter}$
indicates inadequate vascularisation); (3)~the filtration fraction
(percentage of infused inulin or creatinine that appears in the
filtrate per minute), against the native kidney target of $\sim
20\,\%$ at the glomerulus. The claim that the construct ``filters
blood'' should be specified: a single nephron unit in a microfluidic
chip filters at a measurable rate but does not constitute a
clinical kidney substitute. Immune compatibility is not addressable
in vitro; in vivo testing in immunocompetent animals at a scale
matching clinical implantation is the relevant evidence.''
A passing submission asks the three concrete questions and identifies
the difference between a research-grade nephron-on-chip and a
clinical-scale kidney substitute.

\begin{exercise}
A start-up advertises ``cellular regeneration therapy'' that returns
aged tissues to youthful function via injection of stem cells.
Identify two engineering questions the claim does not answer
(homing, persistence, off-target differentiation) and write a
one-paragraph reply to the company.
\end{exercise}

\paragraph{Discussion.} A model reply: ``Thank you for your
information. The claim that injected stem cells return aged tissue
to youthful function reduces to three engineering questions that we
ask you to address: (1)~Homing: what fraction of injected cells
reach the intended tissue, by what mechanism (chemokine-receptor
guidance, mechanical entrapment in microcapillaries, or
deposition on the vascular wall in passage), and over what
timecourse? Existing data for intravenously injected mesenchymal
stem cells (mid-2010s) suggest $< 1\,\%$ of injected cells reach
any specific target tissue and the majority lodge in the lung
microvasculature on first pass; we would like to see your team's
measurement of homing fraction to the claimed target tissues.
(2)~Persistence: how long do the homing cells remain at the target
site, and what fraction of the therapeutic effect is attributable
to the cells themselves versus the paracrine signalling burst that
occurs during the first hours after injection? Many therapies
attributed to cellular regeneration are operationally indistinguishable
from a one-time paracrine bolus and do not require living cells in
the recipient tissue. (3)~Off-target differentiation: in the
fraction of cells that persist, what is the differentiation fate
distribution, and what is the rate of unwanted lineages (teratoma
risk for pluripotent cells, fibrosis risk for mesenchymal cells,
ectopic ossification for cells with osteogenic capacity)? We are
particularly interested in the long-term follow-up data on
unwanted-tissue formation in your treated patients.'' A passing
submission asks the three concrete questions and demonstrates that
the marketing claim is sufficient to imply but not to demonstrate
any of the three.

\begin{exercise}
A reader argues that tissue engineering has failed because no
solid-organ replacements have reached clinical use thirty years
after the field's foundation. Identify two engineering achievements
of the field (skin substitutes, cartilage repair, corneal grafts)
and one engineering reason for the solid-organ stalemate
(vascularisation barrier), and write a one-paragraph reply.
\end{exercise}

\paragraph{Discussion.} A model reply: ``The premise treats clinical
solid-organ replacement as the field's definition of success and is
mistaken on its own terms. Three clinical achievements of tissue
engineering, with regulatory approval and aggregate patient counts
in the hundreds of thousands, are skin substitutes (Apligraf,
Dermagraft, integra-class products for chronic wounds and burn
care; FDA-approved since the late 1990s); autologous chondrocyte
implantation for focal cartilage defects (FDA-approved since 1997,
now generic and reimbursed); and corneal-epithelial-stem-cell
grafts for limbal stem-cell deficiency (European Medicines Agency
approval 2015). Each is a thin-construct application that operates
within the diffusion limit and does not require an engineered
internal vasculature. The solid-organ stalemate is a real
limitation, not a sign of field-wide failure: it is a specific
unsolved problem (engineering an internal vasculature that
anastomoses with the host circulation at clinical scale) on which
the field has made measurable progress (sacrificial-template
fabrication, organoid-bud co-culture, decellularised-matrix re-
endothelialisation) without yet producing a clinical product. The
distinction between ``no solid-organ replacement is clinical yet''
and ``tissue engineering has failed'' is the same distinction
that, in aerospace, separates ``no hypersonic civilian aircraft is
in service yet'' from ``aviation has failed''.'' A passing submission
makes the same two points: thin-construct successes are real and
substantial; the solid-organ barrier is a specific engineering
problem (the vascularisation barrier of this chapter), not a
field-wide failure.

\begin{exercise}
A clinical centre proposes to transplant brain organoids derived from
a patient's iPSCs to replace tissue lost to a stroke. Identify two
engineering questions the proposal does not answer (functional
integration into surviving circuits, scale matching to the lesion,
absence of vasculature in the organoid) and write a one-paragraph
reply naming the most consequential.
\end{exercise}

\paragraph{Discussion.} A model reply: ``The proposal raises three
engineering questions on which we have no answers in the available
materials. (1)~Functional integration: the surviving cortex has
specific axonal projections and synaptic connectivity that the
organoid would need to recapitulate. The cerebral organoid protocol
(Lancaster 2013) produces ventricular-like regions and laminated
cortical layers but lacks the long-range projections, callosal
connections, and thalamic input that would integrate the implant
into a working circuit; the patient would gain tissue mass but not
function unless the integration problem is solved. (2)~Scale: a
stroke lesion is centimetres in linear dimension; a cerebral
organoid is millimetres. Filling a $5\,\text{cm}^3$ cavity with
$3\,\text{mm}$ organoids requires at least one hundred organoids
implanted in parallel, with the diffusion-limited size constraint
applying to each. (3)~Vasculature: the cerebral organoid is
avascular, with the core necrotic and the outer
$\sim 200\,\si{\micro\meter}$ viable. Implantation into a
$5\,\text{cm}^3$ cavity with no vascularisation strategy is
inconsistent with the diffusion-limit budget. The most consequential
of the three is functional integration: even a perfectly
vascularised, perfectly scaled cortical tissue mass that does not
connect to the surviving circuitry will not restore function and
may produce seizure-like activity. We ask the proposing team to
specify the functional-integration measurement (single-unit
recordings, intrinsic-imaging studies, behavioural recovery on
specific motor and cognitive tasks) before clinical use.'' A passing
submission identifies the three questions, names functional
integration as the most consequential, and proposes a measurement.

\begin{exercise}
A vendor advertises a ``universal donor'' tissue construct produced
from gene-edited iPSCs with reduced HLA expression. Identify the
engineering trade-off the design accepts (reduced surveillance against
tumour formation in the implanted cells) and write a one-paragraph
reply asking the vendor to specify the monitoring strategy for that
trade-off.
\end{exercise}

\paragraph{Discussion.} A model reply: ``The ``universal donor''
construct addresses the allogeneic-rejection failure mode by
reducing the donor cells' HLA expression below the threshold for
recipient T-cell recognition. The strategy accepts a specific
engineering trade-off: the same HLA presentation that the
recipient T-cells use to recognise donor cells is also used to
identify cells that have begun malignant transformation. In an
HLA-cloaked construct, transformed cells are invisible to the
recipient's tumour-immune-surveillance system. The
iPSC-derived starting cell population already carries a small but
non-zero rate of integration-associated mutations
and reprogramming-induced genomic instability; in a non-cloaked
construct the recipient's surveillance compensates for this rate,
but the cloaked construct eliminates that compensation. The
monitoring strategy we ask the vendor to specify: (1)~the
pre-implantation genomic-instability screen (whole-genome
sequencing, karyotype, copy-number variation analysis); (2)~the
post-implantation tumour-surveillance protocol (PET or MRI imaging
at three-month intervals for at least three years, with a defined
threshold for explantation); (3)~the suicide-gene safety switch
built into the iPSC genome (e.g., inducible caspase-9 or HSV-TK)
that allows therapeutic elimination of the implanted population
on detection of malignancy; (4)~the long-term recipient-cohort
follow-up plan for tumour incidence at multiples of the natural
background rate. The trade-off is acceptable if the safety-switch
and monitoring protocol detect transformation events before they
become clinically consequential. We are particularly interested in
the recipient enrolment criteria; immunosuppressed patients (for
whom the universal-donor strategy is most attractive) are also
the patients least able to mount a backup surveillance response
against the construct's transformed cells, which is the worst
case for the trade-off.'' A passing submission names the
surveillance trade-off, asks for the safety-switch and monitoring
protocol, and identifies the immunosuppressed-recipient subgroup as
the highest-risk cohort.
