\chapter{Near-misses and weak signals}
\label{vol10:ch13}

\epigraph{Every accident has a near-miss in its past; the discipline is to find the near-miss before the accident finds it.}{}

\chapmeta{Half-life: durable. Archetypes: failure, ethics. Exercise target: 15. Project track: hybrid.}

\noindent
Near-misses are the events that almost happened. The hazard arose; the consequences did not. The cause is sometimes the timely intervention that caught the developing problem, sometimes a fortuitous condition that prevented the consequences, sometimes a mitigation that operated as designed, and sometimes nothing at all (the events simply did not align). Near-misses are the engineering profession's cheap data: they reveal the same conditions that produce accidents, at orders of magnitude greater frequency and without the human cost. The disciplines that systematically gather, analyse, and act on near-misses are among the highest-leverage activities in safety engineering. This chapter is about those disciplines, the signal-recognition skills they depend on, and the cases in which the signals were present and ignored.

\section{Why near-misses are the cheap data}\pathtag{core}

The relationship between near-misses and accidents was framed quantitatively by Heinrich in 1931 (the "Heinrich pyramid" or "accident triangle"); the original ratios he proposed (1 major injury : 29 minor injuries : 300 no-injury incidents) have been criticised, refined, and replaced by more sophisticated models, but the underlying observation has held up: accidents are the tip of an iceberg whose base is many more incidents that did not produce consequences. The base is the cheap data; the tip is the expensive lesson.

The cost asymmetry is the engineering case for near-miss reporting. An accident produces casualties, equipment loss, regulatory action, litigation, reputational damage; a near-miss produces a report. The information content of the report (in terms of identifying the conditions, the mechanism, the contributing factors) is often comparable to the information content of the accident investigation, because the conditions and mechanism are the same. The near-miss is the same event without the consequences.

The conditions under which near-miss reporting actually delivers the value:

\textbf{The reports are filed.} Near-misses are observed by people who are in a position to file the report or not. The default in most organisational cultures is not to file (filing produces work, attention, and possible scrutiny; not filing produces nothing). Filing rates are driven by the perceived utility of filing (does the organisation actually act on the reports?), the perceived risk of filing (will the filer be punished, scrutinised, or otherwise made uncomfortable?), and the ease of filing (is the reporting system fast and accessible?).

\textbf{The reports are analysed.} A pile of near-miss reports without analysis is dead data. The analysis identifies patterns (multiple reports of the same condition; multiple reports involving the same equipment; multiple reports under similar operational conditions) that no single report would surface. Analysis requires capable analysts, dedicated time, and analytic tools that can handle the volume.

\textbf{The patterns are acted on.} The analysis identifies the condition; the action changes the condition. Action requires authority (the analyst's findings reach decision-makers with the budget and authority to act), commitment (the organisation actually performs the changes rather than only acknowledging them), and follow-through (the changes are verified to have addressed the conditions). The Roselawn ATR-72 case \cite{acc:ntsb-aar96-01} is the canonical recent example of analysis without action: multiple prior icing-handling incidents had been reported and analysed; the manufacturer and the regulators had access to the analyses; the action that would have prevented the Roselawn accident was not taken before the accident occurred.

\textbf{The actions feed back into the reporting culture.} When the organisation acts on reports, future filers see the consequence; filing rates rise. When the organisation does not act, future filers see the futility; filing rates fall. The reporting culture is a feedback loop whose stable state is determined by the organisation's commitment to acting on reports.

The engineering economics of near-miss reporting are favourable to the point of obviousness: the marginal cost per report is small; the marginal information value per report is comparable to a small fraction of an accident's lesson; the volume advantage is one or more orders of magnitude. Organisations that do not invest in near-miss systems are leaving the value on the table.

\begin{principle}[Treat near-misses as data]
A near-miss is the same event as an accident, minus the consequences and minus the human cost. The discipline of systematically gathering, analysing, and acting on near-miss data is among the highest-leverage activities in engineering safety. An organisation that does not have a functioning near-miss system is operating without the data its own activity is producing.
\end{principle}

\section{Reporting cultures: aviation ASRS, medical reporting}\pathtag{core}

The Aviation Safety Reporting System (ASRS) operated by NASA on behalf of the FAA is the canonical functioning near-miss reporting system. ASRS receives voluntary reports from pilots, controllers, dispatchers, mechanics, and other aviation professionals; the reports describe events the filer considers safety-relevant. The reports are de-identified (the filer's identity, the operator's identity, and the specific flight identity are removed in the published reports), are protected from regulatory use against the filer for the reported event (with exceptions for criminal activity, intentional violations, and accidents involving fatalities or substantial damage), and are analysed by NASA staff to identify patterns and to publish periodic findings.

The ASRS receives approximately 100,000 reports per year (current as of 2024); the de-identified reports are publicly searchable in the ASRS database. The system's value is established: it has been the source of recommendations that have led to regulatory changes, to procedure changes, to equipment changes; the academic safety-engineering community uses ASRS data extensively for empirical analysis. The system's protections (de-identification, protection from regulatory use) are the structural features that distinguish it from the FAA's Service Difficulty Report system (which is not protected from regulatory action against the operator) and from the NTSB's accident-investigation system (which is invoked only after consequences have occurred).

The medical analogues have had more mixed development. The Patient Safety Organisation (PSO) framework established by the Patient Safety and Quality Improvement Act of 2005 provides federal protection for medical near-miss reports filed through PSOs; the framework's adoption has been substantial but uneven across the healthcare system. The Veterans Health Administration operates an internal patient-safety-event reporting system that is among the most mature medical systems. The aviation analogue's full transfer to medicine has been hampered by the structural differences between aviation (a small number of professional operators, with regulated training and procedures) and medicine (a large number of practitioners with varying institutional structures, regulatory environments, and incentive systems).

The structural conditions for a functioning reporting system:

\textbf{Confidentiality and non-retaliation.} Filers must be confident that filing will not produce personal or institutional consequences. The protection must be substantive (legal protections; institutional commitments; de-identification) rather than only rhetorical.

\textbf{Separation from regulator.} The reporting system must be institutionally separated from the regulator whose enforcement could affect the filer. ASRS's operation by NASA, not by FAA, is the structural realisation of this separation in aviation.

\textbf{Ease of reporting.} The reporting form must be quick to complete; the channels must be accessible; the time cost to the filer must be small relative to the act being reported.

\textbf{Feedback loops.} Filers must see that their reports are acted on (through published findings, through observed changes, through periodic communications). Without feedback, the filing rate degrades to zero.

\textbf{Analytic capacity.} The reports must be analysed by trained personnel with sufficient time and tools. The volume can be substantial; analysis at scale requires investment.

\textbf{Institutional commitment to action.} The findings must lead to changes. Reports that are filed, analysed, and not acted on produce the worst kind of dead data: the appearance of a functioning system without the substance.

\section{Weak-signal recognition}\pathtag{standard}

Weak signals are the indicators that something is wrong with a system, where the indicator's strength is insufficient to compel attention but is sufficient to inform if attended to. The signals are weak by construction: a strong signal is a clear alarm that no one misses; a weak signal is an anomaly that could be normal variation or could be a precursor. The discipline of weak-signal recognition is the discipline of attending to anomalies before their strength rises to the level of consequence.

The mechanisms by which weak signals are missed:

\textbf{The signal is masked by noise.} The anomaly is within the noise envelope of the measurement system; the analyst cannot distinguish it from random variation. The defence is improved measurement (lower noise; longer integration); the cost is the resources required.

\textbf{The signal is masked by frame-shift.} The interpretive frame within which the analyst evaluates the data has shifted in ways that classify the anomaly as normal (Vaughan's normalisation of deviance, Chapter 10). The defence is calibrated re-derivation against the original frame; the cost is the discipline to perform it.

\textbf{The signal is not received.} The data the signal would appear in is not being captured, not being transmitted, not being routed to the analyst. The defence is observability: the deliberate instrumentation of operations to make signals receivable; the cost is the instrumentation investment.

\textbf{The signal is received but not attended to.} The analyst is overloaded; the signal arrives in a queue with many other signals; the queue's processing rate is exceeded by the arrival rate. The defence is prioritisation, queue management, additional analytic capacity; the cost is the operational expense.

\textbf{The signal is attended to but classified incorrectly.} The analyst sees the signal; the analyst's interpretive frame classifies it as benign. The defence is structured classification (with explicit alternative hypotheses), peer review, independent analysis; the cost is the additional analytic effort.

\textbf{The signal is classified correctly but not escalated.} The analyst recognises the signal as concerning; the escalation chain does not transmit the concern to the decision level required. The defence is escalation requirements that are mandatory rather than discretionary; the cost is the organisational discipline.

\textbf{The signal is escalated but not acted on.} The decision-makers receive the concern; the organisational pressures (production, schedule, cost) lead to deferral or rejection. The defence is the structural protections of Chapter 10 (independent technical authority; protected channels for safety concerns; whistleblower mechanisms); the cost is the institutional commitment.

The Columbia accident \cite{acc:caib-columbia} is the canonical case in which signals were received, partly attended to, partly classified incorrectly, and not escalated effectively. Foam-debris strikes during ascent had occurred on multiple prior flights; engineers within NASA had recognised the pattern as concerning; the institutional response had classified the strikes as a turnaround maintenance issue rather than a flight-safety issue. During the Columbia mission, engineers who recognised that the specific strike on STS-107 might have caused critical damage requested on-orbit imagery; the request was declined by the program management. The CAIB identified each of these as a failure of the weak-signal recognition and response chain.

The Lion Air precursor flight to Lion Air 610 \cite{acc:knkt-lion-air-610-2019} is another canonical case: the day before the accident, the same airframe had experienced the same MCAS-driven trim-runaway behaviour on Lion Air Flight 043; an off-duty pilot in the jumpseat had diagnosed and resolved the issue. The event was not escalated as the warning it was; the next day's flight crew did not have the resolution procedure; the accident occurred. The 737 MAX MCAS failure mode had been observed in a controlled (jumpseat-pilot-intervention) form approximately 24 hours before its uncontrolled occurrence killed 189 people \cite{acc:house-737max-2020}.

\section{The HRO posture}\pathtag{core}

High-reliability organisations (HROs) are organisations that operate hazardous technologies for extended periods with substantially fewer accidents than would be predicted by their hazard exposure. The HRO literature (Weick, Sutcliffe, La Porte, Roberts, beginning in the late 1980s) identified the operational dispositions that distinguish HROs from comparable organisations with worse safety records. The HROs include: US Navy carrier flight operations; nuclear power plant operations under INPO discipline; air traffic control; surgical operating rooms operating under high-reliability practices.

The HRO dispositions, abstracted from the empirical literature:

\textbf{Preoccupation with failure.} HROs treat each near-miss as evidence of latent failure conditions; they actively seek out small anomalies that could indicate larger problems. The disposition is the opposite of complacency about safety record: a long accident-free period is read as evidence that the next accident is overdue and that the precursors should be more thoroughly searched for, not as evidence that the system is safe.

\textbf{Reluctance to simplify.} HROs resist the cognitive economy of simplified explanations of complex situations; they treat the complexity as real and the simplification as a shortcut that may hide important features. The disposition resists the pressure to file an incident report into a familiar category and to close it without investigation.

\textbf{Sensitivity to operations.} HROs maintain detailed situational awareness of current operations; the people doing the work are heard by the people managing the work; the operational realities are not filtered out by management's preference for clean reports.

\textbf{Commitment to resilience.} HROs develop the capacity to absorb surprises through training, through redundancy, through rapid response capability, through the institutional flexibility to reconfigure rapidly. The capacity is invested in before it is needed; it is not built up after the first surprise.

\textbf{Deference to expertise.} HROs defer to the person with the relevant expertise regardless of organisational rank in the moment of decision; the authority structure adapts to the situation rather than constraining it. The carrier flight deck during launch and recovery operations is the canonical example: the deck supervisor's authority in that moment exceeds the authority of officers many ranks senior.

These dispositions are not unique to the HROs in the literature; they appear in lesser concentrations in many organisations that do consequential work. The HRO framework is the empirical demonstration that they can be sustained at high concentration in a single organisation, and the analysis of the structural conditions (training, leadership, performance evaluation, resource allocation) that allow the concentration to be sustained.

\begin{archetype}[interface]
Weak-signal recognition is the interface archetype between the system (which generates the signals) and the organisation (which receives them). The interface's properties (the signal-to-noise ratio in observation; the frame within which signals are interpreted; the channels through which they are escalated; the responsiveness of the decision processes) determine whether signals become actions or whether they accumulate into the next accident.
\end{archetype}

\section{Anomaly detection from operational data}\pathtag{standard}

The volume of operational data modern systems generate is large enough that human analysis cannot keep pace with all of it. Anomaly detection is the algorithmic and statistical work of surfacing the data that warrants human attention; the discipline has matured substantially in the era of pervasive instrumentation.

The methods divide along several axes:

\textbf{Threshold-based detection.} A measurement crosses a pre-defined limit; an alert is generated. The simplest method; the appropriate baseline for many applications. The thresholds must be set with attention to false-positive rates (a threshold set too tight produces alarm flood and alarm fatigue) and false-negative rates (a threshold set too loose misses the conditions of concern). Threshold-based detection works well for variables whose normal range is well-characterised and whose anomalous range is clearly distinct.

\textbf{Statistical-process-control detection.} The measurement is monitored against its statistical history (Shewhart charts, CUSUM, EWMA); deviations beyond statistical norms (specified in standard deviations from mean, with explicit Type I error rates) generate alerts. The methods are well-established in manufacturing process control; their adaptation to broader engineering monitoring requires attention to the assumptions (Gaussianity; stationarity) which may not hold.

\textbf{Multivariate detection.} The signal is in the relationship between multiple measurements rather than in any single measurement's deviation. Principal component analysis, multivariate Shewhart, and other multivariate techniques identify deviations from the joint distribution. The methods catch anomalies that univariate methods miss; they require richer characterisation of the normal state.

\textbf{Machine-learning detection.} Algorithms trained on normal-operation data flag deviations from learned patterns. The approaches range from simple (k-nearest-neighbours; isolation forests) to elaborate (autoencoders; deep-learning anomaly detection). The methods can detect complex, multidimensional anomalies; they require substantial training data, careful evaluation of false-positive rates, and explicit handling of concept drift (the "normal" pattern changes over time, and the algorithm must adapt).

\textbf{Semantic detection.} The anomaly is not in the data values but in the meaning of the events: an unusual sequence of operations, an unprecedented combination of conditions, an event that no scenario in the operating envelope anticipated. The detection requires domain knowledge and is often performed by human analysts with algorithmic assistance.

The integration of anomaly detection into operational practice has its own failure modes. Alarm flood: too many alerts overwhelm the human responders, who triage in ways that miss the consequential alerts. Alarm fatigue: chronic false-positive alerts train the responders to dismiss alerts as noise; the consequential alert is missed in the dismissal habit. Algorithmic opacity: the responders do not understand why the algorithm flagged the event, cannot evaluate the flag, and either over-trust (acting on every flag without verification) or under-trust (dismissing flags they cannot interpret). The defence is alarm prioritisation, explicit triage protocols, periodic review of the algorithms' false-positive and false-negative rates, and continued human-analyst capability to evaluate the algorithmic outputs.

\section{Case: organisations that ignored weak signals; ATR-72 icing, Comair 5191}\pathtag{core}

\subsection*{ATR-72 icing: American Eagle 4184, Roselawn, 1994}

The American Eagle 4184 accident on 31 October 1994 killed all 68 occupants when the ATR-72 lost roll control during a hold at 10,000 ft in freezing-drizzle conditions \cite{acc:ntsb-aar96-01}. The mechanism was ice accumulation on the upper wing surface aft of the deicing boots, producing aileron hinge-moment reversal that caused uncommanded roll on autopilot disconnect. The accident's weak-signal story is documented in the NTSB report.

The ATR-72 had been involved in multiple in-service icing incidents in freezing-drizzle conditions in the years before Roselawn. The incidents had not produced fatalities; the crews had recovered control after the anomalous handling; the events had been reported to the manufacturer (the ATR consortium) and to the certification authorities. The reports constituted a weak-signal pattern: the same aircraft type, the same operating environment (cruise in cold-cloud conditions), the same anomalous handling, with multiple independent occurrences. The pattern did not produce flight-manual restrictions, did not produce crew procedures specific to the condition, did not produce design changes to the deicing system, and did not produce a re-evaluation of the icing-certification basis.

The NTSB's investigation identified the failure of the institutional response to the prior incidents as a contributing cause of the Roselawn accident. The signal had been received (the incident reports); the analysis had been performed (the manufacturer and the regulators had examined the reports); the classification had been made (the events were treated as crew-handling issues or one-off events rather than as indications of a certification gap); the escalation that would have prompted the action that would have prevented the accident had not occurred.

The post-accident regulatory response addressed the underlying gap: the FAA issued an immediate Airworthiness Directive prohibiting ATR 42 and ATR 72 operation in known freezing-drizzle conditions; the manufacturer developed extended deicing boots; the FAA revised the icing-certification envelope through the Appendix O process to add the supercooled-large-droplet conditions that the original Appendix C envelope had omitted. The response was substantive; the question the case raises is why the response was conditional on the accident rather than on the prior weak signals.

\subsection*{Comair Flight 5191, Lexington, 2006}

The Comair 5191 accident on 27 August 2006 killed 49 of 50 occupants when the Bombardier CRJ-100 attempted takeoff from the 3500-ft general-aviation runway 26 at Lexington Blue Grass Airport instead of the assigned 7000-ft commercial runway 22 \cite{acc:ntsb-aar07-05}. The aircraft did not achieve flying speed before the runway end and crashed in the terrain beyond. The crew, the single tower controller (who turned to administrative work after issuing the takeoff clearance), the airport's lighting configuration (runway 26 unlit, runway 22 lit, with both visible from the taxi path), and the airport construction project that had altered taxi routing all contributed.

The weak-signal story in Comair 5191 includes the pre-accident hazard assessments at Lexington Blue Grass Airport. The airport's configuration during the construction project had been the subject of safety assessments that identified the potential for runway confusion in the altered taxi routing; the assessments had not led to interim mitigations (improved signage, lighting changes, mandatory crew briefings, or enhanced tower procedures). The signal had been received and analysed; the action had not been taken.

The post-accident response addressed the conditions: FAA changes to tower-staffing requirements (mandating two controllers during operations with commercial flights at certain category airports); FAA changes to runway-incursion-avoidance procedures (mandatory verification of runway alignment against expected heading before takeoff); industry adoption of cockpit-procedure changes that incorporate explicit runway-identification callouts; enhanced crew briefings for airports under construction.

The two cases (Roselawn and Comair 5191) and others in the same pattern (Columbia foam-strike history; Lion Air Flight 043 as precursor to Flight 610; Therac-25's pre-incident anomaly reports; BP Texas City's pre-accident incident pattern) accumulate to a general claim: the weak-signal patterns that precede major accidents are almost always present in the organisation's pre-accident record, and the failure to act on them is the structural condition that distinguishes the organisations whose weak-signal systems work from those whose do not.

\begin{principle}[The signal was probably there]
For a major accident in a high-reliability industry, the prior pattern of incidents, anomalies, and reports almost certainly contains the weak signals that would have predicted the accident. The investigative discipline is to search the pre-accident record for these signals; the operational discipline is to act on signals before the accident occurs. The two disciplines are the same discipline at different times.
\end{principle}

\begin{project}[Design a near-miss reporting protocol]
Select one organisation the reader belongs to or interacts with regularly (a workplace, a software team, a sports club, a household, a research lab, a community group). Design a near-miss reporting protocol for the organisation. The protocol must specify: what counts as a near-miss in the organisation's context; how reports are filed (channel, format, time cost); how reports are protected (confidentiality, non-retaliation); how reports are analysed and by whom; how findings are communicated back to the organisation; how actions resulting from findings are tracked; how the protocol itself is evaluated and revised. Deliverable: a three-page document specifying the protocol, with rationale for each design choice. Include at least one specific example of a hypothetical near-miss in the organisation's context that the protocol would catch.

\textbf{Hazard.} The exercise is not theoretical; if the organisation does not have a near-miss system, the design is a real proposal that the reader may want to advocate for. The professional register and the implementation realism of the proposal matter.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Heinrich's original 1931 framework proposed a ratio of approximately 1 major injury : 29 minor injuries : 300 no-injury incidents. Assume a base population of $10^6$ exposure-hours per year with a major-injury rate of $0.5$ per $10^6$ hours. Compute the expected counts of major injuries, minor injuries, and no-injury incidents per year under the Heinrich ratio. Discuss the predictive value and the limitations of treating the ratio as fixed.
\end{exercise}

\begin{exercise}
The ASRS receives approximately 100,000 reports per year. If each report requires 30 minutes of analyst time on average, compute the analyst headcount required (treating each analyst as 1800 productive hours per year). Compare to plausible NASA staffing for the ASRS analytical function.
\end{exercise}

\begin{exercise}
A statistical-process-control chart uses 3-sigma control limits. For a measurement whose normal distribution has known mean and standard deviation, compute the probability of a false-positive alert per sample. Compute the expected time between false positives for a sampling rate of 1 per minute.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
For a CUSUM control chart with reference value $k$ and decision interval $h$, derive the expected run length (ERL) under the in-control condition. Identify the parameter regime in which the chart has acceptable false-positive rate while retaining sensitivity to small persistent shifts.
\end{exercise}

\begin{exercise}
Derive a model of the reporting-feedback loop: filing rate $f$ depends on perceived utility $U$ and perceived risk $R$; $U$ depends on observed action rate $a$ on prior reports; $a$ depends on $f$ (more reports yield more findings, and so more actions, up to analyst capacity). Identify the stable states of the loop and the conditions that distinguish a functioning reporting culture from a degenerate one.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the ratio of pre-Roselawn ATR-72 icing-handling incidents to the post-Roselawn analysis count \cite{acc:ntsb-aar96-01}. Discuss what the ratio implies about the weak-signal recognition gap in the period.
\end{exercise}

\begin{exercise}
Estimate the cost of operating ASRS (NASA staffing, infrastructure, communication). Compare to the value of one accident prevented by an ASRS-derived recommendation. Identify the ratio at which the system pays for itself.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate a population of 1000 weak-signal observations, of which 5\% are true precursors to accidents and 95\% are benign anomalies. The analyst's classification rule misclassifies 10\% of precursors as benign and 5\% of benign as precursors. Compute the analyst's actual precision and recall. Discuss the implications for action thresholds.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a near-miss reporting system for a hospital surgical unit. Specify the reporting categories, the channels, the protections, the analysis workflow, and the institutional feedback mechanisms. Identify the differences from the aviation ASRS pattern and the reasons.
\end{exercise}

\begin{exercise}
Design an anomaly-detection system for a large cloud-services operation. Specify the data sources, the detection methods, the alert prioritisation, and the human-review workflow. Identify the alarm-fatigue mitigations.
\end{exercise}

\begin{exercise}
Design a weak-signal recognition training module for an engineering team. Specify the cases the training uses (Columbia foam-strike history, Lion Air Flight 043, Roselawn precursors, Comair 5191 pre-accident assessments), the cognitive skills the training builds, and the assessment of training effectiveness.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: an organisation that has had no major accidents in the last five years is more likely than average to have a major accident in the next year, because its weak-signal recognition disciplines have likely degraded with the long accident-free period. Identify the empirical evidence that bears on the argument and the conditions under which it holds or fails.
\end{exercise}

\begin{exercise}
Discuss the structural reasons that the Roselawn ATR-72 icing-incident reports \cite{acc:ntsb-aar96-01} did not produce the regulatory action that the post-Roselawn analysis identified as appropriate. Identify what would have been required in the pre-Roselawn period for the action to have been taken.
\end{exercise}

\begin{exercise}
The Lion Air Flight 043 precursor to Flight 610 is a 24-hour weak-signal case: the same airframe, the same failure mode, the same airline, resolved by a jumpseat pilot's diagnosis \cite{acc:knkt-lion-air-610-2019,acc:house-737max-2020}. Discuss the institutional mechanisms that would have escalated the Flight 043 event to the action that would have prevented the Flight 610 accident.
\end{exercise}

\begin{exercise}
Argue: the engineering profession's standard accident-investigation discipline (NTSB-style retrospective analysis after a major event) is structurally insufficient because it operates only after the consequences have occurred. The weak-signal system (ASRS-style prospective analysis before consequences) is the necessary complement. Discuss what changes in funding, regulatory mandate, and engineering culture would shift the balance toward prospective analysis.
\end{exercise}
