\chapter{Time, frequency, and signals}
\label{vol01:ch06}

\epigraph{The second is defined by taking the fixed numerical value
of the caesium frequency $\Delta\nu_{\text{Cs}}$ to be
$9{,}192{,}631{,}770$ when expressed in the unit
$\si{\hertz}$.}{\textit{The International System of Units}, 9th
edition, BIPM, 2019 \cite[\S 2.3.1]{std:bipm-si2019}.}

\chapmeta{Half-life: foundational for the physics of time, period,
phase, sampling, and aliasing; durable for the institutional
arrangements of UTC and the GNSS constellations; medium-life for the
specific performance numbers of commercial clocks and oscillators.
Archetypes: scaling (time as the quantity engineers measure across
roughly thirty orders of magnitude), interface (the recurring handoff
between a continuous physical signal and a discrete digital record).
Exercise target: 28. Project track: household plus analysis (hybrid),
per Q55-ch06. Hazard class: low. Reader path default: standard, with
sections explicitly tagged. Named cases: Patriot missile system at
Dhahran (1991), as the failure case that closes the chapter; GPS
week-number rollover events (1999 and 2019), as a short-form
companion case for finite-width counter design in time-transfer
chains.}

\noindent
A clock is the first instrument. Engineering inherits its working
notion of time from physics, its working notion of frequency from
the same place, and its working notion of phase from the discipline
of measuring how two periodic signals stand relative to one another.
The reader who finishes the chapter can name the clocks in their
working environment, the conditions under which each is trustworthy,
the sample rate at which a digital recording captures the signal it
claims to capture, and the failure mode by which a small clock error
becomes a large engineering one.

\section{Time as the most precisely measured quantity}\pathtag{core}

Time is the engineering quantity humans measure best. The defining
caesium frequency $\Delta\nu_{\text{Cs}} = 9{,}192{,}631{,}770 \,\si
{\hertz}$ fixes the second exactly, and primary frequency standards
realise the second with fractional uncertainties at the $10^{-16}$
level for cesium fountains and lower for the laboratory optical
clocks that have not yet redefined the SI second
\cite{web:nist-cesium-fountain,web:nist-optical-clock}. Few
engineering quantities are realised with comparable fractional
precision. National mass, temperature, and length realisations can
be extraordinarily good, but their routine uncertainty budgets
depend on method, range, and laboratory. For the purpose of this
chapter, the important fact is narrower: frequency standards reach
fractional uncertainties small enough that many other measurements
are tied back to time or frequency when the measurement chain
allows it.

This precision sets engineering budgets. A second realised at
fractional uncertainty $10^{-16}$ corresponds to gaining or losing
about $3 \,\si{\nano\second}$ per year. A working engineer in 2026
often has access to a disciplined local time signal through GPS,
network time service, or a dedicated oscillator. Time also enters
other measurement chains: length is realised through the fixed
speed of light and measured intervals; Josephson voltage standards
connect voltage to frequency; Kibble balances connect mass to the
Planck constant through electrical and mechanical power. The
practical lesson is that time and frequency often provide the
cleanest reference in the chain, even when the final measurand is
length, voltage, mass, or position.

\begin{archetype}[Scaling, recurring from chapters 1 and 2]
The scaling archetype, introduced in Chapter 1 as the engineer's
internal ruler and developed in Chapter 2 through dimensional
analysis, recurs here on a single quantity. Engineers measure time
across roughly thirty orders of magnitude: from the attosecond
($10^{-18} \,\si{\second}$) light pulses used in atomic-physics
experiments, through the femtosecond and picosecond timescales of
electronic switching, the nanosecond of high-speed digital logic,
the microsecond of GPS travel-time differences, the millisecond of
audio sampling, the second of mains-frequency cycles, the minute and
hour of human work, the day and year of celestial cycles, the
millennium of geological measurement, up to the age of the universe
($\sim 4 \times 10^{17} \,\si{\second}$). Each decade has its
working instruments and its characteristic failure modes. The
engineer's habit is to know which decade a measurement lives in
before reaching for an instrument.
\end{archetype}

\subsection{What a duration is and what an instant is}

A \engterm{duration} is the difference between two instants. An
\engterm{instant} is a label on the time axis. The duration is the
quantity an engineer measures with a stopwatch; the instant is the
quantity an engineer measures with a clock. The two are related by
subtraction, and the uncertainty in a duration measured by two
clocks at different instants involves the uncertainty in each clock
reading and any drift between the readings.

A \engterm{timescale} is a coordinated assignment of instants to a
sequence of physical events, maintained to a stated accuracy and
made available through a publication or a transmitted signal. The
working timescales an engineer in 2026 routinely encounters are
International Atomic Time (TAI), Coordinated Universal Time (UTC),
GPS time, the local civil time of the engineer's jurisdiction, and
the local time produced by an isolated quartz oscillator inside a
piece of equipment. The five timescales differ in offset, in update
rule, and in the institution that maintains them.

\subsection{TAI, UTC, and the leap second}

\engterm{International Atomic Time (TAI)} is the timescale formed
from a weighted ensemble of hundreds of atomic clocks at national
metrology laboratories, computed by the BIPM and reported through
Circular T \cite{web:bipm-circular-t}. TAI runs at the rate of an
SI second realised on the rotating geoid; it has no leap seconds
and no offsets after the initial epoch. TAI is the reference against
which other timescales are compared.

\engterm{Coordinated Universal Time (UTC)} is the timescale used for
civil time worldwide. UTC is produced from TAI by inserting
\engterm{leap seconds}, integer-second adjustments that keep UTC
within $0.9 \,\si{\second}$ of UT1, the timescale derived from the
Earth's rotation. Earth's rotation is slow, irregular, and slightly
slowing on average, so UTC has gained leap seconds whenever the
International Earth Rotation and Reference Systems Service announces
one. As of 2026, UTC is $37 \,\si{\second}$ behind TAI: $10$
seconds at the 1972 establishment of the leap-second mechanism, plus
$27$ leap seconds inserted between 1972 and the most recent
insertion. The last leap second was inserted at the end of December
2016. As of 2026, the mechanism remains live; CGPM Resolution 4 of
2022 directs that the maximum allowed UT1-UTC difference be
increased in or before 2035, which would end the present pattern of
frequent leap-second insertions \cite{web:nist-leap-seconds,
web:bipm-cgpm2022-utc}.

\engterm{GPS time} is the timescale broadcast by the GPS satellite
constellation. Its epoch is 00:00 UTC on 6 January 1980, and it is
not stepped for leap seconds. As of 2026, GPS time is $18 \,\si
{\second}$ ahead of UTC and $19 \,\si{\second}$ behind TAI; the GPS
navigation message carries the GPS-UTC offset so receivers can
present UTC when required \cite{web:usno-gps-time}.

\subsection{Applied vignette: how large systems survive a leap
second}

A leap second is a one-second discontinuity in UTC, and a
discontinuity is precisely what a computer's notion of time is not
built to tolerate. Most operating systems represent civil time as a
count of seconds since an epoch and assume that count increases
monotonically. Inserting a leap second forces one of three
behaviours: repeat the second $23{:}59{:}59$ so the same timestamp
occurs twice, freeze the clock for one second, or step the clock
backward by one second. Each breaks a program that assumes
timestamps strictly increase. The historical record is unforgiving:
the 30 June 2012 leap second triggered lockups in web services,
airline reservation systems, and Linux kernels whose high-resolution
timer code deadlocked when the clock stepped, and several large
operators took hours to recover.

The working engineering response, adopted across the large-scale
operators, is the \engterm{leap smear}: instead of inserting the
extra second at midnight, the time service spreads it across a
window (commonly $24 \,\si{\hour}$, sometimes centred on the
insertion) by running every clock slightly slow. Over a
$24 \,\si{\hour}$ smear window, the fractional rate offset needed to
absorb one second is $1 \,\si{\second} / 86{,}400 \,\si{\second}
\approx 1.16 \times 10^{-5}$, so every server's clock runs about
$11.6 \,\si{\micro\second}$ slow per second for a day, and no
timestamp ever moves backward or repeats. The cost is that a smeared
clock disagrees with UTC by up to half a second during the window,
which is fine for logging and transaction ordering and unacceptable
for a metrology laboratory or a synchrophasor network that needs
true UTC to the microsecond.

The vignette carries the general lesson without any new physics: a
discontinuity in a reference signal is an engineering hazard, and the
mitigation trades a bounded, known rate error for the elimination of
the discontinuity. A reader who designs a system that consumes UTC
should know which behaviour their time source applies at a leap
second, because a smeared source and a stepped source disagree by up
to half a second for a day around the event, and a system that mixes
the two silently misorders records across that window. The
CGPM Resolution 4 of 2022 direction to widen the tolerance on
UT1-UTC by or before 2035 would retire the frequent-leap-second
regime and with it the smear, but until then the smear is the
working practice for any system where monotonic time matters more
than sub-second agreement with UTC \cite{web:nist-leap-seconds}.

\subsection{Why the second went first}

The historical pattern is informative. The metre was redefined in
terms of the second (via the speed of light) in 1983, a half-century
before the kilogram's 2019 redefinition. The reason is the precision
gap: an interferometric length measurement against a wavelength
standard reaches roughly $10^{-9}$ to $10^{-11}$, while a frequency
measurement against a caesium standard reaches $10^{-13}$ to
$10^{-16}$. When a length can be expressed as a time interval times
a fixed velocity, that length inherits the smaller fractional
uncertainty. The same logic applies to mass: when mass can be
expressed through the Planck constant and the Josephson and
quantum-Hall effects, that mass inherits the time precision through
the frequency measurements that ground the constants.

Working principle: reducing a quantity to a time measurement moves
the accuracy floor to the clock and to the model that connects the
clock to the original measurand.

\section{Clocks: pendulum, quartz, atomic, optical}\pathtag{core}

A \engterm{clock} is a physical system whose configuration advances
through repeatable, countable states, together with a counter that
records the number of states traversed. The four working clock
families that span the engineer's vocabulary are the pendulum, the
quartz crystal oscillator, the atomic frequency standard, and the
optical frequency standard. Each operates on a different physical
phenomenon; each has a characteristic accuracy, drift rate, size,
power consumption, and price. We catalogue them as instruments, with
the same discipline applied in Chapter 5 to sensors.

\subsection{The pendulum}

A pendulum's small-amplitude period is approximately
\[
T = 2\pi \sqrt{\frac{\ell}{g}},
\]
where $\ell$ is the effective length and $g$ is the local
gravitational acceleration. The first amplitude correction is
approximately $\theta_{0}^{2}/16$ for swing amplitude $\theta_{0}$
in radians, so a $10^{\circ}$ swing changes the period by about
$0.2\,\%$ and a $23^{\circ}$ swing by about $1\,\%$. A one-metre
pendulum has a period close to $2 \,\si{\second}$, which makes the
seconds pendulum a convenient construction.

Christiaan Huygens designed the pendulum clock in 1656. By the
nineteenth century, regulator pendulum clocks reached fractional
accuracies of $10^{-7}$ to $10^{-8}$ in laboratory conditions: a few
seconds per year. The Shortt-Synchronome free-pendulum clock,
introduced in 1921, was the time standard at major observatories
into the 1940s; it housed the master pendulum in an evacuated tank
to remove air-density variation and used a slave pendulum to do all
the mechanical work, leaving the master pendulum to swing
unimpeded.

The pendulum's accuracy floor is set by sensitivity to local
gravity, temperature (which expands the rod), pressure (which
changes the buoyancy and damping), and amplitude. Each of these is
an \engterm{environmental coupling} that an engineer can either
control or compensate for. Modern pendulum clocks have been
displaced by quartz and atomic standards, but they remain the
classroom example of a time instrument whose physics the reader can
work out from first principles.

\textbf{Worked example.} A steel pendulum has a rod of length $\ell
= 1.000 \,\si{\meter}$ at $20 \,\degC$. The linear thermal-expansion
coefficient of steel is approximately $\alpha = 1.2 \times 10^{-5}
\,\si{\per\kelvin}$. A room-temperature excursion of $5 \,\si
{\kelvin}$ changes the length by $\Delta \ell = \alpha \ell \Delta
T = 6.0 \times 10^{-5} \,\si{\meter}$, a fractional change $\Delta
\ell / \ell = 6.0 \times 10^{-5}$. The period scales as $\sqrt
{\ell}$, so the fractional change in period is half the fractional
change in length, $\Delta T_{\text{p}}/T_{\text{p}} = 3.0 \times
10^{-5}$. Over a $24 \,\si{\hour}$ day this corresponds to a
cumulative timing error of $\Delta T_{\text{p}}/T_{\text{p}} \times
86{,}400 \,\si{\second} \approx 2.6 \,\si{\second}$ per day. A
regulator clock that needs better than $10^{-7}$ fractional
accuracy must therefore either control the rod's temperature
(Shortt-Synchronome's evacuated tank approach) or compensate the
length thermally (the invar-and-brass gridiron pendulum, whose
materials' differential expansion cancels the net change). The
working lesson is that one environmental coupling can dominate the
error budget of an otherwise precise instrument.

\textbf{Worked example: the gridiron compensation ratio.} The
gridiron pendulum cancels the temperature drift of the previous
example by combining two metals whose different expansion
coefficients pull the effective length in opposite directions. Steel
rods (coefficient $\alpha_{\text{s}} = 1.2 \times 10^{-5} \,\si{\per
\kelvin}$) carry the bob downward as they lengthen; brass rods
(coefficient $\alpha_{\text{b}} = 1.9 \times 10^{-5} \,\si{\per
\kelvin}$) push it upward. For zero net change in the effective
length, the heated brass length $\ell_{\text{b}}$ must expand as much
as the heated steel length $\ell_{\text{s}}$:
\[
\alpha_{\text{s}} \ell_{\text{s}} = \alpha_{\text{b}}
\ell_{\text{b}} \quad\Longrightarrow\quad
\frac{\ell_{\text{b}}}{\ell_{\text{s}}} =
\frac{\alpha_{\text{s}}}{\alpha_{\text{b}}} =
\frac{1.2}{1.9} = 0.63 .
\]
For a seconds pendulum whose net effective length is close to
$1 \,\si{\meter}$, the builder therefore proportions roughly
$0.63 \,\si{\meter}$ of brass against $1 \,\si{\meter}$ of steel in
the rod frame. If the ratio is off by $10\,\%$, the residual
fractional length change over a $5 \,\si{\kelvin}$ excursion is about
$0.1 \times \alpha_{\text{s}} \times 5 = 6 \times 10^{-6}$, which
scales to $3 \times 10^{-6}$ in period and $0.26 \,\si{\second}$ per
day, a factor of ten better than the uncompensated steel rod but far
short of what the ratio delivers when it is trimmed correctly. The
working lesson is that compensation converts a first-order
sensitivity into a residual set by the accuracy of a ratio, and the
ratio is a construction tolerance the builder can measure and adjust.

\subsection{The quartz crystal oscillator}

A \engterm{quartz crystal} is a cut sliver of crystalline silicon
dioxide that vibrates at a sharp mechanical resonance when driven
electrically through the piezoelectric effect. The crystal is shaped
and cut to a specific resonant frequency, commonly $32{,}768 \,\si
{\hertz}$ (which equals $2^{15}$ and divides cleanly to a one-second
tick) for wristwatch and timekeeping applications, or megahertz to
hundreds of megahertz for radio and digital electronics.

Wristwatch-grade quartz oscillators are often specified in parts per
million rather than parts in $10^{9}$; $10^{-6}$ corresponds to
about $32 \,\si{\second}$ per year. Temperature-compensated crystal
oscillators (TCXOs) and oven-controlled crystal oscillators (OCXOs)
improve that figure by controlling the crystal's temperature
environment, but the exact value is a datasheet claim, not a
property of quartz in general. The quartz oscillator's error budget
includes aging, temperature, mechanical stress, drive level, and
load capacitance. A quartz reference that has been on a shelf for
ten years should be treated as an oscillator with a history, not as
a fresh datasheet value. The quartz oscillator is the workhorse
frequency reference of consumer electronics, scientific instruments,
and almost every digital system in the reader's environment.

\subsection{The atomic frequency standard}

An \engterm{atomic clock} uses a transition between two well-defined
energy states of an isolated atom as its frequency reference. The
caesium-133 hyperfine transition at $9{,}192{,}631{,}770 \,\si
{\hertz}$ is the SI definition of the second. Working atomic clocks
fall into three classes:

\begin{itemize}[leftmargin=*]
  \item \engterm{Caesium beam clocks}, the workhorse of national
    timing laboratories from the 1960s through the 1990s. A beam
    of caesium atoms passes through a microwave cavity tuned near
    the transition; the resulting atomic state population, detected
    after the beam exits, locks the cavity frequency to the atom.
    Commercial caesium beam standards reach fractional accuracies
    of about $10^{-12}$ to $10^{-13}$.
  \item \engterm{Caesium fountains}, the present primary standards
    in major national laboratories. Laser-cooled caesium atoms are
    launched upward, allowed to traverse a microwave cavity twice
    (once on the way up, once on the way down), and detected. The
    longer interrogation time gives a narrower resonance and
    fractional accuracies near $10^{-16}$
    \cite{web:nist-cesium-fountain}.
  \item \engterm{Chip-scale atomic clocks (CSACs)}, the smallest
    commercial atomic standards available in 2026. They use coherent
    population trapping in a small alkali-vapour cell, fit on a
    circuit board, and draw far less power than rack-scale atomic
    standards. Their stability, drift, aging, power, and price are
    commercial specifications and should be read from current
    datasheets; the durable point is that CSACs trade ultimate
    accuracy for portable holdover time.
\end{itemize}

The atomic clock's accuracy floor is set by systematic shifts of the
transition frequency: collisions between atoms (collisional shift),
gravitational redshift (atoms higher in the Earth's gravitational
field run faster, by about $1.1 \times 10^{-16}$ per metre of
elevation, large enough to require routine correction in primary
standards), magnetic-field-induced shifts, blackbody radiation
shifts, and the quality of the microwave cavity that interrogates
the atom. Each shift is characterised, modelled, and corrected at
fractional levels appropriate to the clock's accuracy class.

\textbf{Worked example.} A caesium fountain operates with a
fractional accuracy near $10^{-16}$. The gravitational redshift
gives a fractional frequency shift of about $1.1 \times 10^{-16}$
per metre of elevation difference between the clock and the geoid
reference. A clock on a $30 \,\si{\meter}$-tall building in a
metrology institute therefore runs fast by about $3.3 \times
10^{-15}$, thirty-three times the accuracy budget of the clock
itself. A standing instruction to record the clock's elevation,
and to apply the redshift correction relative to the geoid, is
therefore a condition on the clock's calibration record, not an
optional refinement. The same reasoning applies to a portable
clock moved between buildings: any vertical displacement of the
clock relocates the fractional-frequency reference.

\Cref{fig:vol01:ch06:allan-deviation} sketches the Allan deviation
$\sigma_y(\tau)$ for four oscillator classes. The Allan deviation
is the standard two-sample statistic for characterising frequency
stability over different averaging times; the short-time
behaviour, the flicker-noise floor, and the long-time drift each
correspond to a recognisable region of the plot. The single-number
fractional accuracy in the table above is the rough centre of the
flicker floor for each family. The Allan deviation lets two
oscillators with the same one-second stability be compared at
one-day or one-year averaging, where they may differ by orders of
magnitude. The script at \texttt{code/allan\_variance.py} computes
the overlapping Allan deviation for a synthetic record.

\input{volumes/01-quantity/ch06-time-frequency-signals/figures/fig-allan-deviation.tex}

\subsection{The optical frequency standard}

An \engterm{optical clock} uses an electronic transition at optical
frequencies (hundreds of terahertz) rather than the microwave
hyperfine transition of caesium (gigahertz). The higher frequency
gives a higher line-quality factor and a narrower resonance,
extending fractional accuracies into the $10^{-18}$ range for the
best laboratory optical clocks
\cite{web:nist-optical-clock}. The bridge between the optical
transition and a frequency that an electronic counter can register
is the \engterm{optical frequency comb}, a laser system whose output
is a coherent set of equally spaced frequency components that span
roughly an octave; the comb provides the gear ratio between the
optical transition (a counted frequency) and the microwave or radio
output that downstream electronics consume.

Optical clocks remain laboratory systems in 2026, with lasers,
vacuum systems, frequency combs, and environmental controls that
keep them outside ordinary field instrumentation. They have not yet
replaced caesium as the SI definition of the second. The CCTF and
national metrology laboratories are preparing the technical basis
for a future optical redefinition; the schedule should be cited to
current BIPM or CCTF material rather than stated as a fixed decade.

\textbf{Worked example: the frequency-comb gear ratio.} An optical
clock ticks at hundreds of terahertz, far above any electronic
counter's reach, so the comb bridges the gap by dividing the optical
frequency down to a countable one. A strontium clock transition sits
near $\nu_{\text{opt}} = 429 \,\si{\tera\hertz}$. A femtosecond comb
with pulse repetition rate $f_{\text{rep}} = 250 \,\si{\mega\hertz}$
produces a spectrum of lines at $\nu_n = f_{\text{ceo}} + n
f_{\text{rep}}$, where $f_{\text{ceo}}$ is the carrier-envelope
offset frequency (a radio-frequency quantity, here taken as $40 \,\si
{\mega\hertz}$). The comb line nearest the clock transition has index
\[
n = \frac{\nu_{\text{opt}} - f_{\text{ceo}}}{f_{\text{rep}}}
= \frac{429 \times 10^{12} - 40 \times 10^{6}}{250 \times 10^{6}}
\approx 1.716 \times 10^{6},
\]
so line number roughly $1.7$ million falls on the transition. The
gear ratio is that index: a fractional change of $10^{-16}$ in the
optical frequency appears as the same fractional change in
$f_{\text{rep}}$, but $f_{\text{rep}}$ is a $250 \,\si{\mega\hertz}$
signal an ordinary counter reads directly. Counting $f_{\text{rep}}$
and $f_{\text{ceo}}$, both radio frequencies locked to the clock,
recovers the optical frequency through the comb equation without ever
counting $429 \,\si{\tera\hertz}$ directly. The working lesson is
that the comb is a frequency divider with a ratio of order a million,
and the two radio-frequency outputs it produces are what downstream
electronics consume; the terahertz transition is never counted, only
divided down.

\subsection{Comparing the four families}

The four clock families form a hierarchy of accuracy and an inverse
hierarchy of cost, size, and accessibility
(Table~\ref{tab:vol01-ch06-clock-families}). The figures are
order-of-magnitude bands for first selection; a real design uses
the current datasheet or metrology report. The reader's working
environment in 2026 contains examples of all four: the pendulum on
the grandfather clock in a museum, the quartz oscillator in every
digital device, the atomic time signal received through GPS or a
network time service, and the optical clocks through their
downstream contribution to UTC.

\begin{table}[h]
\centering
\begin{tabular}{lll}
\toprule
Clock & Fractional accuracy & Working role \\
\midrule
Pendulum (regulator)        & $10^{-7}$ to $10^{-8}$  & Historical;
classroom \\
Quartz wristwatch           & $10^{-6}$               & Consumer
timekeeping \\
Quartz TCXO                 & $10^{-7}$ to $10^{-8}$  & Instrumentation
\\
Quartz OCXO                 & $10^{-9}$ to $10^{-11}$ & Lab reference
\\
Chip-scale atomic           & $10^{-10}$ to $10^{-11}$& Portable atomic
\\
Caesium beam (commercial)   & $10^{-12}$ to $10^{-13}$& National
timing \\
Caesium fountain            & $\sim 10^{-16}$         & SI
realisation \\
Optical (laboratory)        & $\sim 10^{-18}$         & Future SI
candidate \\
\bottomrule
\end{tabular}
\caption{Working clock families and their fractional accuracies as
of 2026. The figures are order-of-magnitude bands suitable for
engineering selection; specific manufacturer datasheets carry the
binding numbers.}
\label{tab:vol01-ch06-clock-families}
\end{table}

\input{volumes/01-quantity/ch06-time-frequency-signals/figures/fig-clock-hierarchy.tex}

\Cref{fig:vol01:ch06:clock-hierarchy} plots the same families on a
common $\log_{10}$ fractional-accuracy axis. The visual span,
twelve decades from the consumer wristwatch to the laboratory
optical clock, is the engineering range over which a designer
chooses a reference. A reader who works exclusively with consumer
electronics rarely needs the right end of that axis; a reader who
works on navigation, telecommunications, or scientific timing
quickly does. The table records the bands; the figure makes the
span legible at a glance. A working version of the table is also
on disk at \repopath{data/clock_families.csv} and feeds the
cumulative-error script at \texttt{code/clock\_comparison.py}.

\subsection{Where each clock fails}

A clock fails when its fractional drift, jitter, or systematic shift
becomes large enough to matter at the application's accuracy
budget. The pendulum fails when the room warms or the air pressure
changes. The quartz oscillator fails when the crystal ages
unattended, when the operating temperature leaves the compensated
range, or when the load capacitance changes after a board repair.
The atomic clock fails when its environmental shielding or its
microwave cavity drifts, or when the operator does not characterise
the systematics required at the clock's accuracy class. The optical
clock fails when its laser stabilisation, its vacuum, or its
frequency-comb link breaks; an optical clock is as much a system as
a sensor.

The engineer choosing a clock should write down the application's
accuracy and stability requirements, the operating environment, the
power and size budget, the calibration record required, and the
cost. The selection follows the same procedure introduced in
Chapter 5 for sensors. A clock is a sensor whose measurand is
elapsed time.

\section{Frequency, period, phase}\pathtag{core}

The three quantities that describe a periodic signal are its
frequency, its period, and its phase. We define each, give its unit,
and state the relationships among them. The reader is asked to keep
the dimensions visible throughout, in the discipline of Chapter 2.

\subsection{Period and frequency}

The \engterm{period} $T$ of a periodic signal is the smallest
positive duration such that the signal repeats: $x(t + T) = x(t)$
for all $t$. Period has dimensions of time, with SI unit the second.

The \engterm{frequency} $f$ of a periodic signal is the number of
periods per unit time:
\[
f = \frac{1}{T}.
\]
Frequency has dimensions of inverse time, with SI unit the hertz,
$\si{\hertz} = 1/\si{\second}$. A signal of frequency $50 \,\si
{\hertz}$ has period $T = 20 \,\si{\milli\second}$; a signal of
frequency $1 \,\si{\giga\hertz}$ has period $T = 1 \,\si{\nano
\second}$. The relationship is exact and runs in both directions.

The \engterm{angular frequency} $\omega$ is the frequency expressed
in radians per second:
\[
\omega = 2\pi f = \frac{2\pi}{T}.
\]
Angular frequency has dimensions of inverse time; the radian is
dimensionless. The factor $2\pi$ is the radian content of one full
period. The reader who confuses $\omega$ with $f$ has dropped a
factor of $2\pi$ and will get sinusoids whose amplitude is right
and whose timing is wrong by that factor.

\subsection{Sinusoids}

A \engterm{sinusoidal signal} of amplitude $A$, frequency $f$, and
\engterm{phase} $\varphi$ is the signal
\[
x(t) = A \cos(2\pi f t + \varphi).
\]
The argument $2\pi f t + \varphi$ is the \engterm{instantaneous
phase}; it advances at the rate $2\pi f$ radians per second. The
phase $\varphi$ at $t = 0$ is the \engterm{initial phase}; it is
the angle by which the signal leads or lags a reference cosine of
the same frequency. Phase is dimensionless; engineering practice
expresses it in radians or degrees.

A periodic signal that is not sinusoidal can be expressed as a sum
of sinusoids at the fundamental frequency $f_{0}$ and its integer
multiples $n f_{0}$, called \engterm{harmonics}. The amplitudes and
phases of the harmonics determine the shape of the signal; this is
the content of Fourier's theorem, which we state informally in
\S 6.5 and which Volume II takes up formally.

\subsection{Phase relationships}

Two sinusoidal signals at the same frequency have a relative phase.
If
\[
x_{1}(t) = A_{1} \cos(2\pi f t + \varphi_{1}), \quad
x_{2}(t) = A_{2} \cos(2\pi f t + \varphi_{2}),
\]
then $x_{2}$ \engterm{leads} $x_{1}$ by $\varphi_{2} - \varphi_{1}$.
A relative phase of $0$ is \engterm{in phase}; a relative phase of
$\pi$ (or $180^{\circ}$) is \engterm{out of phase}; a relative phase
of $\pi/2$ ($90^{\circ}$) is \engterm{in quadrature}. Engineers
working with AC power systems, audio mixing, RF receivers, and
control loops live inside these distinctions every day.

A \engterm{phase shift} can be produced in time. A signal delayed
by $\tau$ seconds at frequency $f$ accumulates a phase lag of
\[
\Delta \varphi = 2\pi f \tau.
\]
At $50 \,\si{\hertz}$ a $1 \,\si{\milli\second}$ delay produces a
phase shift of $0.10\pi$ radians, or $18^{\circ}$. At $1 \,\si{\giga
\hertz}$ the same $1 \,\si{\milli\second}$ delay produces $10^{6}$
full rotations of phase, which is the same as zero phase shift if
all that is observed is the position on the unit circle. The
engineer who measures phase across two signals must know the
unwrapped phase, the cycle count, or both.

\Cref{fig:vol01:ch06:phase-relationships} draws the three canonical
relationships on a common time axis. The reference cosine, the
quadrature signal a quarter period behind it, and the out-of-phase
signal a half period behind it are the offsets an engineer reads off
a two-channel oscilloscope by aligning zero crossings.

\input{volumes/01-quantity/ch06-time-frequency-signals/figures/fig-phase-relationships.tex}

\textbf{Worked example: reading a delay off two oscilloscope
channels.} Two $60 \,\si{\hertz}$ AC waveforms are displayed on a
two-channel scope. Channel B crosses zero going positive $0.92 \,\si
{\milli\second}$ after channel A crosses zero going positive. The
period is $T = 1/60 \approx 16.67 \,\si{\milli\second}$, so the
fractional offset is $0.92/16.67 = 0.0552$ of a period, which is
$\Delta\varphi = 2\pi \times 0.0552 = 0.347 \,\si{\radian} =
19.9^{\circ}$. Channel B lags channel A by about $20^{\circ}$. A
technician who reads this as a power-factor angle infers
$\cos(19.9^{\circ}) = 0.94$, a lagging power factor consistent with
an inductive load. The reverse check: a $20^{\circ}$ lag at $60 \,\si
{\hertz}$ corresponds to a delay $\tau = \Delta\varphi/(2\pi f) =
0.347/(2\pi \times 60) = 0.92 \,\si{\milli\second}$, recovering the
measured offset. The one trap is the half-cycle ambiguity: if the
scope trigger picks the wrong zero crossing, the reading is off by a
full $180^{\circ}$, and only the known sign of the load (inductive
lags, capacitive leads) resolves which crossing is the right one.

\subsection{The mains as a frequency reference}

Most synchronous AC grids operate at a nominal $50 \,\si{\hertz}$ or
$60 \,\si{\hertz}$. The permitted frequency band, the disturbance
limits, and any time-error correction practice are grid-code
matters set by the operator of each synchronous area, not universal
constants. In some synchronous areas the operator also tracks
integrated time error and steers the frequency over longer periods,
so a mains-driven clock can keep long-term time on the grid; in
other areas this practice has been changed or suspended.

In many synchronous grids the mains waveform gives the reader a
measurable frequency signal; where the grid operator publishes a
reference record, the reader can compare a household measurement
against an institutional one. The chapter project asks the reader
to identify the rule for their own synchronous area before
comparing a household record with a published reference.

\subsection{Phase noise and jitter}

A real oscillator does not produce a perfect sinusoid. Its
instantaneous phase carries a random component on top of the
ideal $2\pi f t + \varphi$. This random component is called
\engterm{phase noise} when described in the frequency domain
(spectral density of phase fluctuations) and \engterm{jitter} when
described in the time domain (statistical distribution of zero
crossings or edge timings around their nominal values). Phase noise
limits the resolution of communication receivers, the timing
precision of digital systems, and the noise floor of frequency
measurement. The two descriptions are related by Fourier transform,
which we cover informally in \S 6.5; Volume II gives the formal
account.

A \engterm{jitter specification} on a clock or signal source is a
statistical statement about the spread of edge timings around the
nominal grid. Jitter is reported as RMS jitter (in seconds), as
peak-to-peak jitter, or as integrated phase noise over a stated
frequency band. Each form makes a slightly different claim about
the spread; the engineer reading the spec sheet should know which
quantity is being reported.

\subsection{The dimensions, restated}

A measurement chain that produces a frequency or a phase carries
dimensions through every step. Period: time. Frequency: inverse
time. Angular frequency: inverse time (radians dimensionless).
Phase: dimensionless. Phase noise spectral density: inverse hertz
(when expressed as squared radians per hertz, the radians are
dimensionless and the result has dimensions of inverse frequency).
Jitter: time. The reader who has internalised the discipline of
Chapter 2 will catch a unit error in any of these by inspection.

\section{Sampling and aliasing (Nyquist informally)}\pathtag{core}

A digital measurement of a continuous signal is a sequence of
samples, each tagged with the time at which it was acquired. The
\engterm{sample rate} $f_{s}$ is the number of samples per unit
time, with SI unit the hertz; the corresponding \engterm{sampling
period} $T_{s} = 1/f_{s}$ is the time between consecutive samples.
The reader should know, by inspection, the sample rate of every
digital signal they work with: $44{,}100 \,\si{\hertz}$ or $48{,}000
\,\si{\hertz}$ for CD-quality and professional audio, $96{,}000$ or
$192{,}000 \,\si{\hertz}$ for high-resolution audio, $1$ to $10 \,\si
{\kilo\hertz}$ for industrial control, $1 \,\si{\hertz}$ to $1 \,\si
{\kilo\hertz}$ for laboratory data acquisition, $30$ or $60 \,\si
{\hertz}$ for video.

\subsection{The Nyquist condition}

To recover a sinusoidal signal of frequency $f$ from its samples,
the sample rate must satisfy
\[
f_{s} > 2f.
\]
The frequency $f_{s}/2$ is called the \engterm{Nyquist frequency} or
\engterm{folding frequency}. The condition has a memorable
heuristic: each cycle of the input must produce more than two
samples, otherwise the samples cannot distinguish the signal from
others. The formal statement, the Whittaker-Shannon sampling
theorem, says that a signal whose spectral content lies entirely
below $f_{s}/2$ can be reconstructed exactly from its samples. The
formal proof uses Fourier analysis and is the content of Volume II
Chapter 4 (or wherever calculus arrives in the reader's working
sequence); we state the result informally and motivate it through
the two-samples-per-cycle picture.

To see the heuristic: a signal at exactly $f = f_{s}/2$ produces
samples that alternate between two values per period if the
sampling phase happens to align with the peaks, or that all sit at
zero if the sampling phase aligns with the zero crossings. Either
way, the samples do not uniquely identify the amplitude. A signal
at $f$ slightly above $f_{s}/2$ produces samples that slowly drift
in phase; the resulting sequence is indistinguishable from a sample
of a signal at frequency $f_{s} - f$, which lies below $f_{s}/2$.
This indistinguishability is the central pathology that the
Nyquist condition rules out.

\subsection{Aliasing}

When the Nyquist condition is violated, signals at frequency $f$
above $f_{s}/2$ appear in the sampled record at the
\engterm{alias frequency}
\[
f_{\text{alias}} = | f - n f_{s} |
\]
for integer $n$ chosen so that $f_{\text{alias}}$ falls in the
range $0$ to $f_{s}/2$. The high-frequency component is
\engterm{aliased} into the baseband and cannot be distinguished
from a genuine signal at the alias frequency. Aliasing is a common pathology of digital signal
acquisition because it produces a plausible low-frequency record
from high-frequency content that the sampler failed to exclude.

A worked example: suppose the reader records a $50 \,\si{\hertz}$
mains signal at a sample rate of $30 \,\si{\hertz}$. The Nyquist
frequency is $15 \,\si{\hertz}$, so the $50 \,\si{\hertz}$ signal
violates the condition. With $n = 2$, the alias frequency is
$|50 - 60| = 10 \,\si{\hertz}$. The sampled record will show a
clean sinusoid at $10 \,\si{\hertz}$, and a reader who does not
know the original signal's frequency will report the wrong number.

A second worked example, familiar from cinema: a four-spoke wheel
has the same visual pattern every $90^{\circ}$. Filmed at $24
\,\si{\hertz}$, a wheel that advances $88.5^{\circ}$ per frame is
physically rotating forward at $88.5^{\circ} \times 24 = 2{,}124
\,\si{\degree\per\second}$, or $5.9$ revolutions per second. Because
each frame is only $1.5^{\circ}$ short of the next spoke-pattern
repeat, the sampled image appears to drift backward by $1.5^{\circ}$
per frame, or $36 \,\si{\degree\per\second}$. The eye accepts the
alias because the sampled image has lost the unwrapped rotation
count between frames.

\input{volumes/01-quantity/ch06-time-frequency-signals/figures/fig-sampling-aliasing.tex}

\Cref{fig:vol01:ch06:sampling-aliasing} shows the same pathology in
the simplest one-dimensional case. The top panel samples a
sinusoid at a rate well above twice its frequency; the samples
trace the underlying signal and reconstruction works. The bottom
panel samples the same signal at a rate below twice its frequency;
the samples lie equally on the original signal and on a
lower-frequency alias, and the digital record reports the alias
rather than the original. The script at
\texttt{code/aliasing\_demo.py} verifies the alias relation
numerically for a $7{,}500 \,\si{\hertz}$ sinusoid sampled at $8{,}
000 \,\si{\hertz}$.

\subsection{Anti-aliasing filters}

The discipline that prevents aliasing is the
\engterm{anti-aliasing filter}: an analog low-pass filter placed
before the sampler whose passband cuts off well below the Nyquist
frequency. A first-order filter with corner frequency at half the
Nyquist frequency attenuates by $-6 \,\text{dB}$ per octave above
the corner; a fourth-order filter, by $-24 \,\text{dB}$ per octave;
higher-order filters, more steeply. The engineer chooses the filter
order, corner frequency, and topology to balance attenuation of
out-of-band content, in-band amplitude flatness, and group-delay
distortion.

A working rule for engineering data acquisition: choose the highest
frequency of interest, choose the largest out-of-band signal that
must be rejected, and then choose a sample rate and analog filter
together. Oversampling by a factor of four to ten over the highest
frequency of interest gives room for a filter transition band. The
anti-aliasing filter should pass the signal band with acceptable
amplitude and phase error and provide the required stopband
attenuation by the Nyquist frequency, $f_{s}/2$.

\subsection{Why Nyquist is necessary but not sufficient}

The Nyquist condition, $f_{s} > 2 f_{\text{max}}$, is a statement
about an idealised band-limited signal. A working sampler sees an
input whose spectral content is not strictly band-limited: noise,
interferers, harmonics, and out-of-band signal energy occupy the
frequencies above $f_{\text{max}}$ that the model conveniently
ignores. The anti-aliasing filter is the device that enforces the
band-limit assumption in hardware. Without it, the sample sequence
records the aliased fold of everything above $f_{s}/2$ along with
the band of interest. The engineer who satisfies Nyquist on paper
and skips the anti-alias filter has satisfied the theorem against
a fiction.

The filter's job decomposes into three regions on the frequency
axis: a passband ($0$ to $f_{\text{max}}$) where the signal must
arrive at the sampler with amplitude flat to within an
application-specific ripple budget; a transition band
($f_{\text{max}}$ to $f_{s}/2$) where the filter rolls off; and a
stopband (above $f_{s}/2$) where the filter attenuates by at least
as much as the sampler's dynamic range. A $16$-bit converter has a
dynamic range of about $96 \,\si{\decibel}$; the stopband
attenuation by Nyquist must therefore reach at least that,
otherwise the loudest unfiltered out-of-band component will alias
back at a level comparable to the quantisation floor and
contaminate the working signal.

\subsection{Anti-alias filter design: Butterworth, Bessel, Chebyshev,
elliptic}

Four filter families dominate working anti-alias filter design;
each trades the same three quantities against each other:
sharpness of rolloff, passband flatness, and group-delay distortion
\cite{text:ch06t3-williams-electronic-filter-2006}.

\begin{itemize}[leftmargin=*]
  \item \engterm{Butterworth} filters have maximally flat passband
    amplitude (no ripple) and a monotonic rolloff at $-6n \,\si
    {\decibel}$ per octave for order $n$. The transition band is
    moderate; the group delay is moderately non-flat near the
    corner. The default choice when no other constraint dominates.
  \item \engterm{Bessel} filters have nearly flat group delay
    throughout the passband and into the transition band, at the
    cost of a slower amplitude rolloff than Butterworth at the
    same order. The working choice when preserving the time-domain
    shape of the signal (audio, pulse measurements, control loops)
    matters more than sharp out-of-band rejection.
  \item \engterm{Chebyshev type I} filters trade passband ripple
    (typically $0.1$ to $1 \,\si{\decibel}$) for sharper rolloff
    than Butterworth at the same order. Chebyshev type II inverts
    the trade, putting ripple in the stopband instead of the
    passband; the passband is monotonic and the stopband has
    equal-ripple sidelobes.
  \item \engterm{Elliptic} (Cauer) filters allow ripple in both
    passband and stopband and achieve the sharpest rolloff per
    order of any of the four families. Group delay is highly
    non-flat near the corner. The working choice when the
    transition band must be compressed (deep-space communications,
    narrow-band radar, the analog front end of an oversampled
    converter).
\end{itemize}

The four families form a Pareto frontier: no filter topology
simultaneously achieves Butterworth's passband flatness, Bessel's
flat group delay, and elliptic's sharp rolloff at low order. The
engineer ranks the three competing properties for the application
and chooses the family that respects the binding constraint
\cite{text:ch06t3-williams-electronic-filter-2006}.

\textbf{Worked example: a $44.1 \,\si{\kilo\hertz}$ audio
anti-alias filter.} A CD-quality audio sampler operates at $f_{s}
= 44.1 \,\si{\kilo\hertz}$. The working audio band extends to
$f_{\text{max}} = 20 \,\si{\kilo\hertz}$. The transition band is
$f_{s}/2 - f_{\text{max}} = 22.05 - 20 = 2.05 \,\si{\kilo\hertz}$,
about $0.137$ octaves wide. For a $16$-bit converter the required
stopband attenuation by Nyquist is at least $96 \,\si{\decibel}$.

A Butterworth filter must attenuate by $96 \,\si{\decibel}$ over
$0.137$ octaves; at $6n \,\si{\decibel}$ per octave the required
order is $n = 96 / (6 \times 0.137) \approx 117$. A $117$-th order
analog filter is impractical: component value sensitivity grows
exponentially with order, the group delay distorts the time-domain
pulse response, and the bill of materials blows out. The CD-era
solution was a steep elliptic analog filter of order $7$ to $9$
with $\sim 0.1 \,\si{\decibel}$ passband ripple and $\sim 90 \,\si
{\decibel}$ stopband attenuation, accepting an audible
near-corner pre-ring on transients. The modern solution
oversamples at $128$ or $256$ times the audio rate using a
delta-sigma converter, filters digitally in the sampled domain at
high precision, then decimates to $44.1 \,\si{\kilo\hertz}$. The
analog front-end filter on a modern audio interface is therefore
$4$th- to $6$th-order Bessel or Butterworth with a corner near
$100 \,\si{\kilo\hertz}$, satisfying the oversampler's Nyquist
with comfortable margin and leaving the steep filtering to the
digital domain \cite{text:ch06-oppenheim-schafer-2010}.

An $8$th-order Bessel filter with $-3 \,\si{\decibel}$ corner at
$22 \,\si{\kilo\hertz}$ provides $48 \,\si{\decibel}$ per octave
asymptotic rolloff. At $30 \,\si{\kilo\hertz}$, about $0.45$
octaves above the corner, the attenuation is approximately $48
\times 0.45 \approx 22 \,\si{\decibel}$ of asymptotic plus a
near-corner contribution of $\sim 40 \,\si{\decibel}$, totalling
about $-60 \,\si{\decibel}$ at $30 \,\si{\kilo\hertz}$. The Bessel
choice preserves the time-domain pulse response (group delay flat
to within a few microseconds across the audio band) at the cost of
less aggressive stopband attenuation than an elliptic of
comparable order would deliver.

\textbf{Worked failure case: $44.1 \,\si{\kilo\hertz}$ audio with
no anti-alias filter.} A microphone preamplifier with a switching
power supply emits a $30 \,\si{\kilo\hertz}$ ultrasonic carrier at
$-30 \,\si{\decibel}$ relative to full scale. Sampled at $44.1
\,\si{\kilo\hertz}$ with no anti-alias filter, the alias frequency
is $|30 - 44.1| = 14.1 \,\si{\kilo\hertz}$, squarely inside the
audible band. The recording contains a tonal artefact at $14.1
\,\si{\kilo\hertz}$ at $-30 \,\si{\decibel}$, $66 \,\si{\decibel}$
above the $16$-bit quantisation floor, audible as a high-pitched
whine. The engineering response is to add a $4$th-order Butterworth
low-pass filter with $-3 \,\si{\decibel}$ corner at $20 \,\si{\kilo
\hertz}$ between the preamplifier and the sampler; at $30 \,\si
{\kilo\hertz}$ the filter attenuates by approximately $24 \times
\log_{2}(30/20) = 24 \times 0.585 \approx 14 \,\si{\decibel}$ of
out-of-band rejection, dropping the alias to $-44 \,\si{\decibel}$
but not eliminating it. A steeper $8$th-order filter (Bessel or
elliptic) brings the alias well below the audible threshold for
the same input level. The lesson: filter order is set by the
loudest expected out-of-band signal and the required suppression
at Nyquist, not by an aesthetic preference for steep rolloff.

\textbf{Worked failure case: a data logger and a pump.} An Onset
HOBO U30 environmental logger samples a structural-vibration
accelerometer at $1 \,\si{\hertz}$ at a wastewater pumping station.
The logger has no analog anti-alias filter; its front end passes
DC to a few tens of hertz. A nearby centrifugal pump runs at
$360 \,\si{rpm} = 6 \,\si{\hertz}$, coupling mechanical vibration
into the building structure. Sampled at $1 \,\si{\hertz}$, the
$6 \,\si{\hertz}$ vibration aliases to $|6 - 6 \times 1| = 0 \,\si
{\hertz}$ (DC). The recorded trace shows a quasi-static offset
over the duration of pump operation, indistinguishable on the trace
from a genuine slow drift of the structure. The engineering review
returns two options: either sample at $\geq 12 \,\si{\hertz}$ to
leave the pump's $6 \,\si{\hertz}$ signature unaliased, or add a
$1$st-order RC anti-alias filter with corner at $0.3 \,\si{\hertz}$,
which attenuates $6 \,\si{\hertz}$ by $\log_{2}(6/0.3) \times 6 =
4.32 \times 6 \approx 26 \,\si{\decibel}$, dropping the alias to a
manageable level. The discipline is the same as the audio case: a
sample rate without an anti-alias filter has not satisfied Nyquist
on the spectral content actually present.

\textbf{Worked failure case: vibration monitoring on an industrial
blower.} A maintenance engineer instruments a $60 \,\si{\hertz}$
shaft-speed centrifugal blower in a refinery utilities plant with a
piezoelectric accelerometer (PCB Piezotronics 352C33,
$100 \,\si{\milli\volt}/g$) and a portable data logger. The
specification calls for harmonic-content monitoring through the
fifth harmonic at $300 \,\si{\hertz}$. The engineer chooses a
sample rate of $100 \,\si{\hertz}$, reasoning that the fundamental
is well inside the Nyquist range. The accelerometer's bandwidth is
flat from $1 \,\si{\hertz}$ to $10 \,\si{\kilo\hertz}$; no
anti-alias filter sits between the accelerometer and the logger's
$24$-bit converter.

The third harmonic at $180 \,\si{\hertz}$ aliases to
$|180 - 2 \times 100| = 20 \,\si{\hertz}$. The fifth harmonic at
$300 \,\si{\hertz}$ aliases to $|300 - 3 \times 100| = 0 \,\si
{\hertz}$ (DC). The spectrum displayed by the logger's analysis
software shows a strong tone near $20 \,\si{\hertz}$ that does not
correspond to any expected mechanical signature of the blower or
its motor. The engineer's first hypothesis, drawn from the
characteristic-frequency table for the blower's $6309$ deep-groove
ball bearing (ball-pass frequency outer race
$\approx 0.4 \times N_{b} \times f_{\text{shaft}}$ for $N_{b} = 8$
balls, giving $\sim 192 \,\si{\hertz}$ at $60 \,\si{\hertz}$ shaft
speed), does not match. The engineer instead reports a
$20 \,\si{\hertz}$ bearing-fault signature, schedules an
inspection, and the maintenance crew removes a healthy bearing.

The disambiguating test: raise the sample rate to $1{,}000 \,\si
{\hertz}$ with a $4$th-order Butterworth anti-alias filter at
$400 \,\si{\hertz}$ corner. The $20 \,\si{\hertz}$ tone disappears
and the spectrum shows clean lines at $60$, $120$, $180$,
$240$, $300 \,\si{\hertz}$, the expected harmonic content of a
healthy non-sinusoidal vibration source. The lesson is the same
one the audio and data-logger cases carry: sample rate without
anti-aliasing budgets nothing against aliased content above
Nyquist. The cost of the wrong diagnosis here is a bearing
replacement, a production outage during inspection, and erosion of
trust in the vibration-monitoring program. ISO 10816-3:2009 sets
the vibration-severity bands against which the spectrum is
interpreted; an aliased spectrum interpreted against ISO 10816 is
a category error.

\textbf{Worked failure case: power-quality monitor on a 5th
harmonic.} A facilities engineer deploys a Fluke 1748 three-phase
power-quality logger at a $480 \,\si{\volt}$ industrial bus
feeding a variable-frequency-drive (VFD) load that injects a
strong fifth harmonic ($250 \,\si{\hertz}$ on a $50 \,\si{\hertz}$
mains) into the supply. The logger is configured to sample voltage
at $f_{s} = 200 \,\si{\hertz}$, a rate chosen to fit a long-
duration recording onto local storage. The front-end anti-alias
filter is a $1$st-order RC at $100 \,\si{\hertz}$ corner,
attenuating $250 \,\si{\hertz}$ by only $20 \log_{10}(250/100)
\approx 8 \,\si{\decibel}$.

The $250 \,\si{\hertz}$ component aliases to
$|250 - 2 \times 200| = 150 \,\si{\hertz}$, which is the third
harmonic of the mains and a known signature of saturated
transformers and arc-furnace loads. The engineer's report
attributes the third-harmonic content to transformer saturation
upstream of the logger, recommends a tap-changer adjustment, and
the utility opens a ticket. The actual mechanism is the VFD's
fifth harmonic aliasing through an under-specified power-quality
logger. The fix is the same shape: raise the sampler to
$f_{s} \geq 1{,}000 \,\si{\hertz}$ (IEC 61000-4-7 specifies
$200 \times 50 = 10{,}000 \,\si{\hertz}$ as the reference sampling
rate for harmonic analysis through the $40$th order on a
$50 \,\si{\hertz}$ system), or replace the front-end filter with a
$4$th-order Butterworth at $100 \,\si{\hertz}$ that attenuates
$250 \,\si{\hertz}$ by $\sim 32 \,\si{\decibel}$. The IEC
$200 \cdot f_{\text{line}}$ rule for harmonic analysers is the
discipline that holds Nyquist at the spectral content the standard
expects to measure; an installer who configures a slower sampling
rate has departed from the standard's assumed budget.

\subsection{Filter order, group delay, and the rolloff budget}

A working anti-alias filter design carries an explicit error
budget across three domains. Passband amplitude ripple ($\Delta A$,
in $\si{\decibel}$) is the maximum permitted variation in passband
gain across $0$ to $f_{\text{max}}$; for audio, $\Delta A \leq
0.1 \,\si{\decibel}$ is the working target. Stopband attenuation
($A_{\text{stop}}$, in $\si{\decibel}$) is the minimum suppression
required above $f_{s}/2$, set by the converter's dynamic range and
the loudest expected out-of-band signal. Group-delay variation
($\Delta \tau_{g}$, in $\si{\second}$) is the maximum permitted
variation in group delay across the passband; for audio,
$\Delta \tau_{g} \leq 100 \,\si{\micro\second}$ across the
passband is the working target, since group-delay variations on
this order audibly distort transient sounds.

The four filter families occupy distinct regions of this
three-dimensional budget space. Butterworth and Bessel give
monotonic passbands ($\Delta A = 0$) and trade rolloff sharpness
against group-delay flatness. Chebyshev and elliptic accept
passband or stopband ripple in exchange for sharper rolloff. The
working procedure is to plot the three required tolerances against
filter order for each family and pick the lowest-order filter
that fits inside the budget; the design references at
\cite{text:ch06t3-williams-electronic-filter-2006} carry the
formulas.

\subsection{Quantisation}

A digital sample is also quantised: the continuous voltage at the
sampling instant is rounded to one of $2^{N}$ levels, where $N$ is
the bit depth of the analog-to-digital converter. The
\engterm{quantisation step} is
\[
q = \frac{V_{\text{range}}}{2^{N}},
\]
where $V_{\text{range}}$ is the full-scale range of the converter.
A $16$-bit converter with $V_{\text{range}} = 5 \,\si{\volt}$ has
$q \approx 76 \,\si{\micro\volt}$; a $12$-bit converter with the
same range has $q \approx 1.2 \,\si{\milli\volt}$.

The quantisation introduces a noise component whose RMS value, for
inputs that exercise many levels, is approximately
$q / \sqrt{12}$, distributed roughly uniformly over the band $0$ to
$f_{s}/2$. The peak-to-peak amplitude of the quantisation error is
$q$. The engineer who reports a sampled measurement should know
both $q$ and the implied quantisation uncertainty, and should add
the latter in quadrature to the other uncertainty components per
the rules of Chapter 4.

\begin{estimation}
\textbf{Estimate the sample rate and bit depth required to record
the chapter project's mains-frequency signal at the required
accuracy.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The mains signal is approximately sinusoidal at $50 \,\si{\hertz}$
(or $60 \,\si{\hertz}$ in some jurisdictions). The reader is asked
to detect the zero crossings of this signal to within enough
precision that the inferred frequency over a one-minute averaging
window has a standard uncertainty of order $1 \,\si{\milli\hertz}$
or finer.

The fractional accuracy required is $1 \,\si{\milli\hertz} / 50 \,\si
{\hertz} = 2 \times 10^{-5}$. Over $60 \,\si{\second}$ the signal
completes $3{,}000$ cycles; a relative timing precision of $2 \times
10^{-5}$ over $60 \,\si{\second}$ corresponds to a timing precision
of $1.2 \,\si{\milli\second}$ on the start-and-end edges of the
window. That is generous: a sample rate of $1 \,\si{\kilo\hertz}$
gives a sample period of $1 \,\si{\milli\second}$, comparable to the
timing budget, and a sample rate of $10 \,\si{\kilo\hertz}$ leaves
margin.

The project therefore chooses a sample rate above the formal
Nyquist minimum because zero-crossing timing, sample-clock quality,
and front-end delay set the measurement budget. With $f_{s} = 10
\,\si{\kilo\hertz}$, each sample is $100 \,\si{\micro\second}$.
Linear interpolation between sign-change samples can improve the
crossing estimate when the waveform is clean and the front-end
delay is stable. Over a $60 \,\si{\second}$ window, the remaining
uncertainty should be checked against the sample clock's fractional
frequency error rather than treated as a property of the sampling
theorem alone.

The bit depth is set by signal-to-noise considerations. A
$10$-bit ADC with $5 \,\si{\volt}$ range gives $q \approx 5 \,\si
{\milli\volt}$; this is comfortably above the noise floor of a
typical audio-input front end and is sufficient to localise zero
crossings to within a single sample. A $16$-bit ADC, common in
modern sound cards, gives $q \approx 76 \,\si{\micro\volt}$, well
beyond the requirement. The reader does not need a precision
laboratory ADC for the project; an ordinary smartphone or laptop
audio input is enough, provided the input stage is AC-coupled and
the signal does not clip the converter range.

The estimate, then, is: $f_{s} = 10 \,\si{\kilo\hertz}$ or higher,
$N = 16$ bits, with the dominant uncertainty falling on
sample-clock quality (which is the laptop's own oscillator) and on
the optical or audio front-end group delay rather than on the
sampling theorem itself.
\end{estimation}

\section{Time-frequency duality preview}\pathtag{standard}

A signal can be described in the time domain (its waveform $x(t)$)
or in the frequency domain (its spectral content $X(f)$). The two
descriptions carry the same information: each can be derived from
the other by a transform that Volume II Chapter 4 will state
formally. We sketch the working content here without invoking
calculus, in the discipline established for this volume.

\subsection{The Fourier picture, informally}

A periodic signal of period $T_{0}$ and fundamental frequency
$f_{0} = 1/T_{0}$ can be written as a sum of sinusoids at the
harmonic frequencies $n f_{0}$ for integer $n \geq 0$:
\[
x(t) = \sum_{n=0}^{\infty} a_{n} \cos(2\pi n f_{0} t + \varphi_{n}).
\]
The amplitude $a_{n}$ and phase $\varphi_{n}$ of each harmonic
together specify the signal; the set of pairs $(a_{n}, \varphi_{n})$
is the signal's \engterm{Fourier representation}. The reverse
direction, recovering the harmonics from the waveform, is what the
formal Fourier transform achieves; for a periodic signal it gives
the Fourier series, for a non-periodic signal of finite duration it
gives the discrete or continuous Fourier transform, depending on
whether the signal is sampled or continuous.

A square wave at fundamental frequency $f_{0}$ has Fourier content
at $f_{0}$, $3 f_{0}$, $5 f_{0}$, $7 f_{0}$, with amplitudes that
fall as $1/n$. A triangle wave has the same odd harmonics with
amplitudes that fall as $1/n^{2}$. A pure sinusoid has content only
at $f_{0}$. The shape of the waveform is the harmonic content; the
harmonic content is the shape of the waveform.

\subsection{The discrete Fourier transform, informally}

A digital recording is a finite sequence of samples,
$x[0], x[1], \ldots, x[N-1]$, taken at sample rate $f_{s}$ over a
duration $N T_{s}$. The \engterm{discrete Fourier transform (DFT)}
turns this finite sample sequence into a finite frequency
representation: a sequence of $N$ complex coefficients
$X[0], X[1], \ldots, X[N-1]$, where $X[k]$ is the complex amplitude
of the spectral component at frequency $k f_{s} / N$. The frequency
spacing of the DFT is $\Delta f = f_{s}/N = 1/(N T_{s})$, which
equals the inverse of the recording duration. For real-valued
records, bins above $N/2$ represent the negative-frequency side of
the spectrum, mirrored in the usual one-sided magnitude plot.

The \engterm{fast Fourier transform (FFT)} is an algorithm that
computes the DFT in time $\mathcal{O}(N \log N)$ rather than the
$\mathcal{O}(N^{2})$ of direct computation. It is the workhorse
algorithm of digital signal processing and the reason that
spectral analysis is a routine engineering operation rather than a
specialist's craft.

\subsection{Time-frequency duality}

The duration of a recording and the frequency resolution of its
spectrum are reciprocals. A one-second recording has a frequency
resolution of $1 \,\si{\hertz}$; a millisecond recording has a
frequency resolution of $1 \,\si{\kilo\hertz}$. To resolve two
spectral peaks separated by $\Delta f$, the recording must last at
least $1/\Delta f$. To localise a feature to within a duration
$\Delta t$, the spectral content must extend over a bandwidth of
order $1/\Delta t$. This is the engineering form of an uncertainty
relation: time and frequency are duals, and a signal cannot be made
arbitrarily short and arbitrarily narrow-band at the same time.

\input{volumes/01-quantity/ch06-time-frequency-signals/figures/fig-time-frequency-duality.tex}

\Cref{fig:vol01:ch06:time-frequency-duality} illustrates the
duality for a rectangular pulse: a long pulse transforms to a
narrow sinc-shaped spectrum; a short pulse transforms to a broad
one. The product $\Delta t \cdot \Delta f$ has a lower bound of
order unity whose precise value depends on the definitions of $\Delta t$ and $\Delta f$ used. Engineering choices about pulse duration,
chirp bandwidth, and matched-filter design all live inside this
trade.

The duality has working consequences. A spectrum computed from a
short recording has poor frequency resolution; a spectrum computed
from a long recording cannot localise transient events. The
\engterm{short-time Fourier transform} and its relatives (the
spectrogram, the wavelet transform) are constructions designed to
trade time and frequency resolution against each other; we
describe them only by name here. Volume VII (Information) takes up
the working theory.

\subsection{Spectral leakage}

A practical DFT is computed on a finite-duration record. If the
record contains a non-integer number of cycles of an input
sinusoid, the DFT spreads the sinusoid's energy across multiple
frequency bins instead of concentrating it in a single bin. This
spreading is called \engterm{spectral leakage}; it is an artefact
of the truncation of the signal to the recording window, not a
property of the underlying signal. Engineering practice mitigates
leakage by applying a \engterm{window function} (Hann, Hamming,
Blackman, flat-top, others) to the time-domain samples before
computing the DFT; the window tapers the record to zero at the
edges and reduces the leakage at the cost of slightly broader
main-lobe response. Window selection is a working trade-off; the
reader who computes a spectrum without a window has implicitly
chosen the rectangular window, the worst choice for most
applications.

\subsection{Decomposing periodic signals informally}

The Fourier picture of \S 6.5.1 gives the working content without
the integral machinery. Three canonical shapes illustrate the
correspondence between waveform geometry and harmonic content.

A pure sinusoid at frequency $f_{0}$ has one component: an impulse
at $f_{0}$ in the magnitude spectrum, no other harmonic content.
The spectrum is the simplest possible: one tone, one bin.

A square wave at fundamental frequency $f_{0}$ has odd-harmonic
content at $f_{0}, 3 f_{0}, 5 f_{0}, 7 f_{0}, \ldots$ with
amplitudes falling as $1/n$. The reader who plots the magnitude
spectrum of a $100 \,\si{\hertz}$ square wave sees discrete spikes
at $100, 300, 500, 700, \ldots \,\si{\hertz}$ with relative
heights $1, 1/3, 1/5, 1/7, \ldots$. A triangle wave at the same
fundamental has the same odd-harmonic locations with amplitudes
$1/n^{2}$: the spectrum decays faster, consistent with the
smoother time-domain waveform. A sawtooth wave has all integer
harmonics with amplitudes $1/n$.

The general working rule: the smoother the waveform (more
continuous derivatives), the faster the harmonic amplitudes fall
with $n$. A perfectly smooth (infinitely differentiable) periodic
signal has a spectrum that decays faster than any power of $1/n$;
a signal with a single discontinuity has spectrum falling as
$1/n$; a signal whose first derivative is discontinuous has
spectrum falling as $1/n^{2}$. This is the working dictionary
between time-domain regularity and frequency-domain compactness;
the formal statement appears in Volume II.

\textbf{Trumpet note worked example.} A trumpet playing a sustained
A4 at $440 \,\si{\hertz}$ produces a complex periodic signal whose
spectrum contains the fundamental at $440 \,\si{\hertz}$ and
strong integer harmonics at $880, 1320, 1760, 2200 \,\si{\hertz}$
and higher. The relative strengths of the harmonics determine the
\engterm{timbre}, the perceived tone colour that distinguishes a
trumpet from a clarinet (which has strong odd harmonics and weak
even ones, the spectral signature of an open-closed pipe) or a
violin (which has many harmonics with a slow decay). The trumpet's
brightness corresponds to a slow harmonic decay rate; the
clarinet's hollow tone corresponds to its odd-only structure. The
spectrum carries the perceptual identity of the instrument; the
fundamental alone carries only the pitch.

\Cref{fig:vol01:ch06:harmonic-content} collects the three canonical
waveforms on one amplitude axis. The sawtooth carries every integer
harmonic at amplitude $1/n$; the square and triangle carry only odd
harmonics, because both are odd-symmetric about their half-period;
the triangle's $1/n^2$ decay outruns the square's $1/n$, the
frequency-domain reading of its smoother time-domain shape.

\input{volumes/01-quantity/ch06-time-frequency-signals/figures/fig-harmonic-content.tex}

\textbf{Worked example: total harmonic distortion of a clipped
sinusoid.} A power amplifier driven past its rails clips the peaks
of a $1 \,\si{\kilo\hertz}$ sinusoid, and the clipped output measures
a fundamental at $1.00 \,\text{V}$ RMS with odd harmonics at
$0.09 \,\text{V}$ ($3 \,\si{\kilo\hertz}$), $0.05 \,\text{V}$
($5 \,\si{\kilo\hertz}$), and $0.03 \,\text{V}$ ($7 \,\si{\kilo
\hertz}$), even harmonics negligible. The total harmonic distortion
is
\[
\mathrm{THD} = \frac{\sqrt{V_3^2 + V_5^2 + V_7^2}}{V_1}
= \frac{\sqrt{0.09^2 + 0.05^2 + 0.03^2}}{1.00}
= \frac{\sqrt{0.0115}}{1.00} = 0.107,
\]
or about $10.7\,\%$. The odd-only, slowly-falling harmonic pattern is
the spectral signature of symmetric clipping: the clipped waveform
approaches a square wave, whose harmonics fall as $1/n$, so the fifth
harmonic at roughly $1/5$ of the fundamental is the working diagnostic
that the amplifier is clipping rather than adding low-order distortion
from a mildly non-linear transfer curve (which would show a stronger
second harmonic). A reader who sees only the fundamental and misses
the harmonic structure reports a clean tone and a distorted amplifier
as the same measurement.

\subsection{Spectrograms: time-localised frequency content}

A spectrum computed once over a long record averages frequency
content over the entire duration. A musical note that changes pitch
during the recording loses its time structure in this single
average: the spectrum shows a smear, not a trajectory. The
\engterm{spectrogram} restores time information by computing many
short FFTs in overlapping windows, plotting frequency on one axis,
time on the other, and amplitude (or power) as colour. The result
is a three-dimensional view of the signal: time, frequency, and
power, simultaneously.

The \engterm{short-time Fourier transform (STFT)} is the
construction underneath. The signal is multiplied by a sliding
window $w(t)$ (typically Hann or Hamming) and the DFT is computed
on the windowed segment. The window slides by a hop fraction of
its width (typically $25\%$ to $75\%$ overlap), producing a column
of the spectrogram every hop step.

\textbf{Worked spectrogram example: violin glissando.} A one-second
recording of a violin player sliding from A4 ($440 \,\si{\hertz}$)
to A5 ($880 \,\si{\hertz}$) at $44{,}100 \,\si{\hertz}$ sample
rate produces a spectrogram with a visible diagonal ridge from
$440$ to $880 \,\si{\hertz}$, plus parallel ridges for each
harmonic (at $2f, 3f, 4f, \ldots$) that rise in tandem.

The window length controls the trade between time and frequency
resolution. A Hann window of $4 \,\si{\milli\second}$ ($176$
samples) gives time resolution $\sim 4 \,\si{\milli\second}$ and
frequency resolution $\sim 250 \,\si{\hertz}$: the ridge appears
sharp in time but the harmonics smear in frequency, individually
unresolvable. A Hann window of $40 \,\si{\milli\second}$ ($1{,}764$
samples) gives time resolution $\sim 40 \,\si{\milli\second}$ and
frequency resolution $\sim 25 \,\si{\hertz}$: the harmonics
separate cleanly but the rising glissando smears in time, looking
like several constant tones at different frequencies during the
window. A working choice for instrumental music is a $20$ to
$30 \,\si{\milli\second}$ window: short enough to track tempo,
long enough to resolve harmonics within a semitone.

\subsection{Time-frequency uncertainty}

The spectrogram exposes a fundamental property of time-frequency
representations: a signal cannot be both arbitrarily short in
time and arbitrarily narrow in frequency. Dennis Gabor's 1946
formulation \cite{text:ch06t3-gabor-1946} of this principle states
that for any reasonable definition of time width $\Delta T$ and
frequency width $\Delta f$,
\[
\Delta T \cdot \Delta f \geq \frac{1}{4\pi},
\]
with equality achieved by a Gaussian-windowed signal. Engineering
practice often uses the looser working form
\[
\Delta T \cdot \Delta f \gtrsim \frac{1}{2},
\]
appropriate when $\Delta T$ and $\Delta f$ are defined as the
full-width at half-maximum of a Gaussian or the equivalent
rectangular width. The constant depends on the definitions; the
product is bounded from below regardless.

\textbf{Worked uncertainty case.} A reader wants to resolve a $1
\,\si{\milli\second}$ transient signal (a switching spike, a click)
in a $1 \,\si{\hertz}$ frequency bin. The product $\Delta T \cdot
\Delta f = 10^{-3} \times 1 = 10^{-3}$ falls below the Gabor
bound $1/(4\pi) \approx 0.0796$. No time-frequency representation
can deliver both resolutions simultaneously. The engineer's
options are to relax one of the two requirements: either accept
$\sim 100 \,\si{\hertz}$ frequency resolution and keep the $1
\,\si{\milli\second}$ time resolution (a short window), or accept
$\sim 0.5 \,\si{\second}$ time smearing and keep $1 \,\si{\hertz}$
frequency resolution (a long window). The choice is the
application's, not the analyst's.

The same physics constrains every time-frequency tool: STFT,
wavelet transform, Wigner-Ville distribution, reassignment
methods. Each redistributes the time-frequency information
differently within the Gabor bound; none escapes it. The
quantum-mechanical Heisenberg uncertainty between position and
momentum is the same mathematical statement applied to the
position-frequency duality of the wavefunction.

\subsection{Where time-frequency analysis matters in practice}

Three working instruments make heavy use of time-frequency
representations.

\engterm{Lock-in detection} extracts a sinusoidal signal at a known
frequency from a noisy background by multiplying the input by a
local reference at the target frequency and low-pass filtering.
The output is the in-phase and quadrature amplitude of the input
at the reference frequency. Stanford Research Systems' SR830 and
SR860 lock-in amplifiers (current as of 2026) reach noise floors
of $\sim 10 \,\si{\nano\volt}/\sqrt{\si{\hertz}}$ across audio and
RF bands, with time constants from microseconds to hundreds of
seconds. The instrument is the working tool for measuring small
periodic signals in optics (chopped beam intensity), thermometry
(lock-in calorimetry), and impedance spectroscopy
\cite{text:ch06t3-pozar-microwave-2011}. A lock-in amplifier is a
narrow-band tracking spectrogram with a single frequency bin.

\engterm{FFT on a digital oscilloscope} brings spectrum analysis
to the bench. The Keysight DSOX1102G entry-level oscilloscope
(current as of 2026) provides an FFT math function with up to
$1$ million points, window choices (Hann, Hamming, Blackman-Harris,
flat-top, rectangular), and a $0.5 \,\si{\milli\hertz}$ frequency
resolution at the longest available record length. The bench
engineer uses the FFT function to identify the dominant tones in a
signal, to measure harmonic distortion of an amplifier, and to
locate the source of an interfering carrier. The same instrument
with the FFT off shows a periodic ringing; with the FFT on, the
ringing resolves into discrete tones whose frequencies match
identifiable physical sources (the switching frequency of a
nearby power supply, the mains harmonic, the clock of a digital
device).

\engterm{Rotating-machinery vibration diagnostics} relies on
spectrograms to track the frequency of mechanical vibration as
the machine's rotational speed changes. A bearing fault appears as
a tone at a specific multiple of the shaft rotation frequency
(the bearing's characteristic frequency, set by ball count and
geometry); during a startup or coastdown, the fault tone tracks
the rotational frequency in a spectrogram and is easily separated
from constant background tones. Working tools (Bruel \& Kjaer
PULSE, National Instruments LabVIEW Sound and Vibration Toolkit)
display the spectrogram in real time and overlay the rotational
frequency from a tachometer signal. Bearing-fault diagnostics in
wind turbines, jet engines, and turbomachinery rest on this
display. The chapter project's mains-frequency record can be
analysed in the same way: a spectrogram of the $50 \,\si{\hertz}$
mains over $24 \,\si{\hour}$ shows the slow drift of the
fundamental, the slow modulation of the harmonics, and any
transient events superimposed on the steady-state signal.

\subsection{When the duality preview is enough}

The chapter project does not require formal Fourier analysis. It
requires the reader to estimate the average frequency of a slowly
varying near-$50 \,\si{\hertz}$ or near-$60 \,\si{\hertz}$ signal in
one-minute windows. Time-domain zero-crossing detection is the
cleanest route: detect sign changes in the sampled waveform and
interpolate between sign-change samples to refine the crossing
times. A frequency-domain route can also work, but a one-minute
FFT has bin spacing $1/60 \,\si{\hertz} \approx 16.7 \,\si{\milli
\hertz}$, so millihertz resolution requires peak interpolation,
phase tracking, or a fitted sinusoid rather than reporting the
largest raw FFT bin. The chapter project asks the reader to
characterise the chosen method's uncertainty and to defend it.

Later chapters take up the machinery this preview names: integration
and series in Volume II, physical waves in the force and energy
volumes, and digital signal processing in Volume VII Chapter 15.

\section{GPS as a measurement system}\pathtag{standard}

The Global Positioning System (GPS), and the broader
\engterm{Global Navigation Satellite Systems (GNSS)} that include
Galileo, GLONASS, and BeiDou, are timing systems that solve for
position through time-of-flight measurements. Navigation is the
visible service; time transfer is the measurement mechanism
underneath it. We describe GPS as a measurement system: clocks,
signals, propagation models, ephemeris, and the geodetic frame in
which the result is expressed.

\input{volumes/01-quantity/ch06-time-frequency-signals/figures/fig-gps-architecture.tex}

\Cref{fig:vol01:ch06:gps-architecture} maps the chain. Each
satellite carries on-board atomic clocks steered by the master
control station against GPS system time, which is itself steered
to UTC(USNO). The receiver records the time of arrival of each
satellite's signal against its local oscillator, forms a
pseudorange, and solves four unknowns from four pseudoranges. The
figure carries the working-number callout that organises the rest
of this section: $1 \,\si{\nano\second}$ of timing error on a
pseudorange is $30 \,\si{\centi\meter}$ on the ground.

\begin{archetype}[Interface, recurring from chapters 3 and 5]
The interface archetype recurs in GPS at multiple levels. The
satellite's clock interfaces to its onboard frequency standard; the
satellite's signal interfaces to the receiver's antenna; the
receiver's local clock interfaces to the GPS system time; the
position solution interfaces to the geodetic reference frame; the
geodetic reference frame interfaces to the user's local map. Each
interface is a place where a measurement crosses from one
representation into another, and each carries its own uncertainty,
calibration history, and failure mode. A GPS measurement that does
not account for every interface in the chain is a GPS measurement
whose uncertainty has not been characterised.
\end{archetype}

\subsection{The constellation geometry}

The GPS space segment is a constellation of nominally $24$
satellites distributed across $6$ orbital planes inclined at
$55^{\circ}$ to the equator, with $4$ satellites per plane plus
spares. The orbital period is one-half sidereal day ($\approx 11
\,\si{\hour}\,58\,\si{\minute}$), so each satellite traces the same
ground track each sidereal day. The orbit altitude is approximately
$20{,}200 \,\si{\kilo\meter}$ above mean sea level, placing the
satellites in medium Earth orbit
\cite{text:ch06t3-kaplan-hegarty-2017}. As of 2026 the active
constellation typically holds $31$ to $32$ satellites, well above
the $24$ minimum needed to guarantee global coverage.

The orbit-plane and inclination design has a working consequence:
from any point on Earth's surface, at any time, at least four
satellites are visible above a $5^{\circ}$ elevation mask in
open-sky conditions, and typically eight to twelve are visible
above $10^{\circ}$. Four is the minimum for a navigation fix; the
remaining satellites improve the geometry and provide redundancy
against signal blockage.

The \engterm{geometric dilution of precision (GDOP)} quantifies the
geometry's effect on position uncertainty. For a pseudorange
uncertainty $\sigma_{\rho}$ shared across all visible satellites,
the position uncertainty is $\sigma_{p} = \mathrm{GDOP} \times
\sigma_{\rho}$. GDOP factors decompose into horizontal (HDOP),
vertical (VDOP), and time (TDOP) components. For an open-sky
four-satellite fix with one satellite near zenith and three
spread around the horizon, HDOP $\approx 1.5$ and VDOP $\approx
2.5$. For a tracking solution with eight to twelve visible
satellites, HDOP drops to $0.7$ to $1.0$ and VDOP to $1.0$ to
$1.5$. In an urban canyon where buildings block the lower
elevations, HDOP may rise to $5$ to $10$ and VDOP higher still,
inflating the horizontal position uncertainty by the same factor
\cite{text:ch06-misra-enge-2011}.

\textbf{Worked geometry example.} A receiver in an open field sees
nine satellites with HDOP $0.9$ and pseudorange uncertainty $\sigma
_{\rho} = 2 \,\si{\meter}$ after atmospheric corrections. The
horizontal position uncertainty is $0.9 \times 2 = 1.8 \,\si
{\meter}$. The same receiver moves to a street between $30 \,\si
{\meter}$ buildings; only four satellites remain visible above the
elevation mask, HDOP rises to $4$, and the same pseudorange
uncertainty inflates to $8 \,\si{\meter}$ horizontally. The
receiver's reported position accuracy follows the geometry, not
the underlying timing precision; the satellites and the receiver
chain have not changed, only the visible subset has.

\subsection{Solving for position: the four-pseudorange fix}

A GPS receiver records, for each tracked satellite $i$, a
pseudorange $\rho_{i}$ defined as the speed of light times the
measured travel time from satellite to receiver. The pseudorange
relates to the geometric satellite-receiver distance and to the
receiver clock bias $b$ (measured in seconds) through
\[
\rho_{i} = \sqrt{(x_{s,i} - x)^{2} + (y_{s,i} - y)^{2} + (z_{s,i} - z)^{2}}
+ c\,b + \epsilon_{i},
\]
where $(x_{s,i}, y_{s,i}, z_{s,i})$ are the satellite's known
ECEF coordinates at signal transmission, $(x, y, z)$ are the
unknown receiver coordinates, $c$ is the speed of light, and
$\epsilon_{i}$ lumps the residual error contributions (atmosphere,
multipath, receiver noise). With four or more pseudoranges, the
four unknowns $(x, y, z, b)$ are over-determined and the receiver
solves them by iterated least squares.

The standard procedure linearises around an estimated receiver
position. Given a starting estimate $(x_{0}, y_{0}, z_{0}, b_{0})$,
the receiver computes the predicted pseudoranges $\hat\rho_{i}$
and the residuals $\Delta \rho_{i} = \rho_{i} - \hat\rho_{i}$. The
Jacobian of the pseudoranges with respect to the four unknowns
has rows
\[
\mathbf{H}_{i} = \begin{bmatrix}
-e_{x,i} & -e_{y,i} & -e_{z,i} & c
\end{bmatrix},
\]
where $\mathbf{e}_{i} = (e_{x,i}, e_{y,i}, e_{z,i})$ is the unit
vector from satellite $i$ toward the estimated receiver position.
The position update is the least-squares solution
\[
\Delta \mathbf{x} = (\mathbf{H}^{\top} \mathbf{H})^{-1}
\mathbf{H}^{\top} \Delta \boldsymbol\rho.
\]
The iteration repeats until $\|\Delta \mathbf{x}\|$ falls below a
tolerance (typically $< 1 \,\si{\milli\meter}$). Three to five
iterations suffice for any reasonable starting estimate; a cold-
start receiver bootstraps from a coarse position derived from the
broadcast almanac or a network-supplied estimate
\cite{text:ch06-misra-enge-2011}.

The clock unknown $b$ collapses the four-pseudorange problem to a
position-plus-time fix at every receiver epoch. The receiver does
not need an accurate local clock; the four-satellite solution
recovers the bias and applies it. This is the central reason GPS
receivers can be built around quartz oscillators with $10^{-6}$
fractional accuracy and still deliver $30 \,\si{\centi\meter}$
ranging precision: the position solution removes the bias every
epoch.

A receiver that loses sight of all but three satellites cannot
solve for the four unknowns. A receiver with a high-quality
holdover clock can fix on three satellites for a brief interval
(the clock bias is treated as known from the last full fix); this
is the working principle of \engterm{aided GPS} during partial
satellite occultation in urban canyons and indoor environments.

\subsection{The GPS clocks}

Each GPS satellite carries four atomic frequency standards:
two rubidium and two caesium-beam clocks in the legacy Block IIR
satellites, and four rubidium clocks in the modernised Block
IIF and Block III spacecraft. The Block IIR rubidium clocks
(manufactured by EG\&G Frequency Products, now Spectratime) achieve
Allan deviation $\sigma_{y}(\tau = 1\,\si{\day}) \approx 1 \times
10^{-14}$ when steered; the caesium-beam clocks (manufactured by
Frequency Electronics) reach $\sigma_{y}(\tau = 1\,\si{\day})
\approx 8 \times 10^{-15}$ but with shorter operational lifetime.
The active clock is selected by ground command from the GPS
Operations Center at Schriever Space Force Base in Colorado
Springs and is monitored continuously by the 17 monitor stations
of the worldwide control segment (NGA monitor stations plus the
6 Air Force monitor stations).

The on-board clock is steered against the GPS system time, a
weighted ensemble timescale defined by the master control station
from the monitor-station measurements. The steering correction
takes the form of broadcast clock-correction polynomial
coefficients $(a_{f0}, a_{f1}, a_{f2})$ in the navigation message
subframe 1, applied by the receiver as
$\Delta t_{\text{sv}} = a_{f0} + a_{f1}(t - t_{oc}) + a_{f2}(t - t_{oc})^{2}$
where $t_{oc}$ is the reference time of the clock data; the
quadratic term accommodates a frequency-drift rate, and the
linear term accommodates a frequency offset. The control segment
uploads new clock-correction coefficients to each satellite about
once per day; the broadcast values represent the master control
station's best prediction of the on-board clock's behaviour
forward to the next upload. The post-fit residual of this
prediction (the actual versus predicted clock state at the time
of upload) is the Signal-in-Space User Range Error contribution
from the clock, typically $\sim 0.6\,\si{\meter}$ at the
$1\sigma$ level for the modernised constellation.

The GPS system time differs from UTC by an integer number of
seconds (the leap-second offset, $18\,\si{\second}$ as of 2026)
plus a small offset typically held within $\pm 50\,\si{\nano
\second}$ ($1\sigma$). GPS does not insert leap seconds; it
broadcasts the current UTC-GPS offset, the time and date of any
scheduled future leap second, and a 6-bit signed week count of
the current leap-second-delta change in subframe 4 page 18 of
the navigation message. Receivers can use these fields to present
civil time correctly when asked. The Galileo, GLONASS, and BeiDou
systems each define their own system times that differ from GPS
by tens to hundreds of nanoseconds; multi-constellation receivers
estimate the inter-system biases as additional unknowns.

\subsection{The ranging measurement}

A GPS receiver measures the time of arrival of a satellite signal
relative to its own local clock. Multiplied by the speed of light,
this travel time gives a \engterm{pseudorange}: the distance from
satellite to receiver, biased by the offset between the receiver
clock and the satellite clock. With four or more satellites in
view, the receiver solves four unknowns (three position
coordinates plus the receiver clock offset) from four pseudoranges.
The position solution is a byproduct of the clock synchronisation;
the clock synchronisation is a byproduct of the position solution.

The fundamental equation behind the ranging is straightforward:
distance equals speed times time. The speed is the speed of light
in vacuum, $c = 299{,}792{,}458 \,\si{\meter\per\second}$ exactly
(as fixed by the SI definition of the metre). A timing error of
$1 \,\si{\nano\second}$ in the receiver's measurement of the
arrival time therefore corresponds to a ranging error of about
$30 \,\si{\centi\meter}$. The reader should retain this as a
working number.

\subsection{Error sources}

The position uncertainty of a single-frequency GPS receiver in
2026 is dominated by a small set of error sources, each with its
own physics and its own engineering response.

\begin{itemize}[leftmargin=*]
  \item \engterm{Ionospheric delay}. The signal traverses the
    ionosphere, a region of free electrons whose density peaks
    near $300\,\si{\kilo\meter}$ altitude and depends on solar
    activity (the $11$-year solar cycle drives a factor-of-three
    variation in total electron content), latitude (equatorial and
    polar anomalies), local time (peak around $14$h local time,
    minimum near sunrise), and season. The first-order delay
    scales as $40.3 \cdot \text{TEC}/f^{2}\,\si{\meter}$ where
    $\text{TEC}$ is the total electron content along the slant
    path in TEC units ($10^{16}\,\text{electrons}/\si{\meter
    \squared}$) and $f$ is the carrier frequency. At L1
    ($1575.42\,\si{\mega\hertz}$), a typical mid-latitude daytime
    TEC of $30\,\text{TECU}$ produces a $\sim 5\,\si{\meter}$
    range error; during ionospheric storms it can exceed
    $50\,\si{\meter}$. The Klobuchar broadcast model corrects
    $\sim 50\%$ of the daily-averaged error for single-frequency
    L1 receivers; the Galileo NeQuick model reaches $\sim 70\%$.
    Dual-frequency receivers cancel the first-order term to
    millimetre precision by forming the ionosphere-free combination
    of L1 ($1575.42\,\si{\mega\hertz}$) and L2 ($1227.60\,\si{
    \mega\hertz}$) or L1 and L5 ($1176.45\,\si{\mega\hertz}$);
    the residual second-order term is $\sim 1\%$ of the first
    order.
  \item \engterm{Tropospheric delay}. The neutral atmosphere
    contributes a zenith path delay around $2.3$-$2.6\,\si{\meter}$
    (about $8$-$9\,\si{\nano\second}$), of which $90\%$ is the
    nearly-stable hydrostatic component (predicted to $\sim 1\,\si{
    \milli\meter}$ from surface pressure via the Saastamoinen
    model) and $10\%$ is the highly variable wet component (from
    water vapour, poorly predicted by surface meteorology). The
    slant delay at elevation $\epsilon$ uses a mapping function
    such as the Global Mapping Function: $\text{md}(\epsilon) =
    1/(\sin\epsilon + a/(b + c\sin\epsilon))$ where $a, b, c$ are
    site-specific coefficients near $0.001$, $0.01$, $0.05$. At
    $5\,\si{\degree}$ elevation, the slant delay is $\sim 25\,\si{
    \meter}$; below $10\,\si{\degree}$, the wet-component
    uncertainty rises rapidly and most receivers apply an
    elevation mask.
  \item \engterm{Multipath}. Reflected signals interfere
    constructively or destructively with the direct signal at the
    receiver antenna. The maximum multipath error for code (C/A)
    measurements is $\sim 0.5 \cdot \tau \cdot c$ where $\tau$ is
    the chip period ($1\,\si{\micro\second}/1023 \approx
    977\,\si{\nano\second}$ for C/A, hence $\sim 150\,\si{\meter}$
    maximum theoretical error), but practical receivers use
    narrow-correlator or strobe-correlator architectures (NovAtel
    PAC, Trimble Maxwell) that limit the error to $\sim 1$-$3\,\si{
    \meter}$ in typical environments and $5$-$50\,\si{\meter}$ in
    deep urban canyons. Carrier-phase multipath is bounded by
    $\lambda/4$ ($\sim 5\,\si{\centi\meter}$ for L1), much smaller
    but still significant for RTK applications.
  \item \engterm{Satellite clock and ephemeris error}. The
    broadcast ephemeris is valid for $\pm 2\,\si{\hour}$ around the
    reference time and is updated every $2\,\si{\hour}$. The
    broadcast User Range Error (URE) is $\sim 0.6\,\si{\meter}$
    for the modernised constellation. The IGS (International GNSS
    Service) publishes Ultra-Rapid (predicted, available real-
    time, $\sim 5\,\si{\centi\meter}$ orbit, $3\,\si{\nano\second}$
    clock), Rapid ($\sim 17$-$41\,\si{\hour}$ latency,
    $2.5\,\si{\centi\meter}$ orbit, $75\,\si{\pico\second}$ clock),
    and Final products ($12$-$18\,\si{\day}$ latency,
    $\sim 2.5\,\si{\centi\meter}$ orbit, $75\,\si{\pico\second}$
    clock) that improve on the broadcast for post-processing
    \cite{web:gpsgov-accuracy}.
  \item \engterm{Receiver noise and observable choice}. C/A code
    measurement noise at $1\sigma$ is $\sim 1\,\si{\meter}$ for
    consumer receivers and $\sim 0.1\,\si{\meter}$ for survey-
    grade receivers; P(Y) code noise is $\sim 0.3\,\si{\meter}$
    and $0.03\,\si{\meter}$ respectively. Carrier-phase
    measurements have $\sim 1\,\si{\milli\meter}$ noise at L1
    ($\lambda = 19.0\,\si{\centi\meter}$) but the integer cycle
    ambiguity must be resolved through dual-frequency observation
    or extended observation time. Survey-grade receivers
    (Trimble R12i, Leica GS18, Sokkia GRX3) reach $\sim 8\,\si{
    \milli\meter} + 1\,\si{ppm}$ horizontal in RTK mode at
    baselines under $30\,\si{\kilo\meter}$.
\end{itemize}

User accuracy depends on satellite geometry, signal-in-space
error, atmosphere, multipath, receiver design, and processing.
GPS.gov reports open-sky smartphone accuracy on the order of a few
metres and a government signal-in-space commitment that must not
be confused with receiver accuracy. Differential GPS (DGPS) and
real-time kinematic (RTK) methods use reference-station
measurements and carrier-phase processing to move from metre-scale
positioning toward sub-metre or centimetre-scale positioning when
the environment, baseline, and ambiguity resolution support it
\cite{web:gpsgov-accuracy,std:gps-sps-2020}.

\subsection{The working RTK and PPP pipelines}

Carrier-phase positioning requires correction streams from
reference stations. Two delivery architectures dominate the
working RTK and PPP markets.

\engterm{Single-baseline RTK} uses one reference receiver at a
surveyed location streaming corrections in real time to the rover.
The reference and rover form a baseline; the rover differences its
own observations against the reference's to cancel the
satellite-clock, ephemeris, ionospheric, and tropospheric errors
common to both endpoints. The accuracy floor scales linearly with
baseline length: at the working Trimble R12i / Leica GS18 receivers,
RTK accuracy is $8\,\si{\milli\meter} + 1\,\si{ppm}$ horizontal,
or $10\,\si{\milli\meter}$ at a $2\,\si{\kilo\meter}$ baseline and
$38\,\si{\milli\meter}$ at $30\,\si{\kilo\meter}$. The
maximum reliable single-baseline RTK distance is about $30$-$40
\,\si{\kilo\meter}$, beyond which decorrelation of the atmospheric
error overwhelms the differencing benefit.

\engterm{Network RTK (NRTK)} replaces the single reference with a
network of $5$-$30$ continuously operating reference stations
(CORS) covering a region; the rover receives a virtual reference
station (VRS) correction stream computed from the network's
observations interpolated to the rover's location. The interpolation
uses ionospheric and tropospheric models tuned from the network's
own data. Working NRTK services: in the US, the National Geodetic
Survey's CORS network ($\sim 2000$ stations) plus commercial
providers Trimble VRS Now, Leica SmartNet, NovAtel TerraStar; in
Europe, the SAPOS network in Germany (one station per $\sim
50\,\si{\kilo\meter}$) and the EUREF Permanent Network for
geodetic-quality data; in Asia, GEONET Japan ($\sim 1300$
stations); in Australia, AUSCORS. NRTK reduces the baseline-length
dependence: working accuracy is $8\,\si{\milli\meter} + 0.5\,
\si{ppm}$ at distances up to $\sim 50\,\si{\kilo\meter}$ from the
nearest network station.

\engterm{NTRIP (Networked Transport of RTCM via Internet
Protocol)} is the de facto correction-stream transport standard,
RTCM Special Committee 104 maintained, current version 2.0 (2011).
The protocol runs over HTTP; a caster (server) accepts correction
streams from sources (reference stations) and delivers them to
clients (rovers) authenticated by username/password. A rover's
NTRIP client sends a $\mathrm{GGA}$ NMEA sentence (its
approximate position) to request a VRS stream from the caster; the
caster returns the interpolated correction stream as RTCM 3.3 or
RTCM 3.4 messages over the persistent HTTP connection. Working
NTRIP casters: NTRIP Caster (BKG, German), MountPoint (Trimble),
ICESec (Leica), and open-source SNIP and BNC (BKG). Rover client
software: BNC, SNIP, the rtklib open-source toolkit by Tomoji
Takasu (BSD licence, the working reference implementation),
commercial NRTK clients in survey controllers (Trimble Access,
Leica Captivate).

\engterm{Precise Point Positioning (PPP)} is the alternative to
RTK: instead of differencing against a nearby reference, the
rover uses precise satellite orbit and clock products from a
global computation centre (the IGS combined Final Products at
$2.5\,\si{\centi\meter}$ orbit and $75\,\si{\pico\second}$ clock;
the IGS Ultra-Rapid products real-time at $5\,\si{\centi\meter}$
and $3\,\si{\nano\second}$). PPP eliminates the need for a nearby
reference at the cost of $15$-$30\,\si{\minute}$ initial
convergence time and slightly larger steady-state uncertainty.
Working PPP services: NRCan CSRS-PPP (free, post-processing),
JPL APPS (free, post-processing), Trimble RTX (commercial,
real-time, $\sim 4\,\si{\centi\meter}$ horizontal after
$5\,\si{\minute}$ convergence), Veripos APEX (commercial,
real-time, offshore-oil-and-gas industry), Sapcorda SAPA
(commercial, real-time, automotive).

\engterm{PPP-RTK} (also called State-Space Representation, SSR)
combines PPP's global reach with RTK's fast convergence by
broadcasting State-Space corrections (decomposed into per-
satellite orbit, clock, code-bias, and atmospheric terms) rather
than raw observation corrections. The Galileo HAS (High Accuracy
Service, operational since 2023) and the Japanese MADOCA service
broadcast SSR corrections over the L6 signal of QZSS and
Galileo's E6-B signal. Working PPP-RTK convergence: $1$-$5\,\si{
\minute}$ versus $15$-$30\,\si{\minute}$ for classical PPP.

The working choice across these architectures is set by the
application's accuracy requirement, geographic extent, real-time
versus post-processing constraint, and budget. A surveyor uses
NRTK in a region with good network coverage; a geodetic
researcher uses post-processed PPP with IGS Final products; an
oil-and-gas dynamic-positioning vessel uses real-time PPP with
Veripos; an autonomous vehicle uses PPP-RTK over Galileo HAS.
The architectures share the carrier-phase ambiguity-resolution
problem and differ in how they generate the corrections that
constrain the integer ambiguity to a known value.

\subsection{Relativistic effects}

Two relativistic corrections are routinely applied in GPS. First,
the satellite clocks run faster than ground clocks because they
are higher in the Earth's gravitational potential, by about
$45 \,\si{\micro\second}$ per day. Second, they run slower than
ground clocks because they move at orbital speed, by about
$7 \,\si{\micro\second}$ per day. The net effect is a satellite
clock that runs faster than a ground clock by about
$38 \,\si{\micro\second}$ per day. The satellite frequency standard
is offset before launch so that the broadcast frequency, after
relativistic effects, matches the GPS system rate. A second-order
correction, periodic over the satellite orbit and of magnitude
tens of nanoseconds, is applied in the receiver firmware. A reader
who turns off the relativistic correction in a receiver simulation
will see position errors growing at roughly $c \times 38 \,\si
{\micro\second\per day} \approx 11 \,\si{\kilo\meter}$ per day.
Relativity sits inside the GPS operating budget
\cite{std:gps-sps-2020}.

\textbf{The two terms, derived.} The fractional clock-rate
difference between a satellite clock and a clock on the rotating
geoid follows from the weak-field metric used in special and
general relativity. The general-relativity term is the
gravitational redshift,
\[
\left( \frac{\Delta f}{f} \right)_{\text{GR}}
= \frac{\Phi_{\text{sat}} - \Phi_{\text{geoid}}}{c^{2}},
\]
where $\Phi$ is the Newtonian gravitational potential ($\Phi \to
-GM/r$ for a point mass). For a GPS satellite at orbital radius
$r_{s} = R_{E} + h \approx 26{,}600 \,\si{\kilo\meter}$ and a
ground clock at $r_{g} = R_{E} \approx 6{,}378 \,\si{\kilo\meter}$,
\[
\frac{\Phi_{\text{sat}} - \Phi_{\text{geoid}}}{c^{2}}
= \frac{GM}{c^{2}} \left( \frac{1}{r_{g}} - \frac{1}{r_{s}} \right)
\approx 5.28 \times 10^{-10}.
\]
The fractional rate excess multiplied by one day ($86{,}400 \,\si
{\second}$) gives a daily offset of $+45.6 \,\si{\micro\second}$:
the satellite clock runs faster.

The special-relativity term is the velocity-dilation slowdown,
\[
\left( \frac{\Delta f}{f} \right)_{\text{SR}}
= -\frac{v_{\text{sat}}^{2} - v_{\text{geoid}}^{2}}{2 c^{2}}.
\]
The satellite's orbital speed is $v_{\text{sat}} \approx 3.87 \,\si
{\kilo\meter\per\second}$; the ground clock's tangential speed at
the equator is $v_{\text{geoid}} \approx 0.47 \,\si{\kilo\meter
\per\second}$. The fractional rate is
\[
-\frac{(3870)^{2} - (470)^{2}}{2 \times (3 \times 10^{8})^{2}}
\approx -8.25 \times 10^{-11}.
\]
Multiplied by one day, this gives $-7.1 \,\si{\micro\second}$: the
moving satellite clock runs slower.

The sum is $+38.5 \,\si{\micro\second}$ per day, the net offset
applied to the satellite clock before launch. The pre-launch
correction is implemented in hardware: the satellite oscillator's
nominal frequency is set to $10.229999995453 \,\si{\mega\hertz}$
rather than $10.23 \,\si{\mega\hertz}$, the fractional offset
$-4.46 \times 10^{-10}$ exactly cancelling the on-orbit net rate
shift. The broadcast carrier, derived from the corrected
oscillator, arrives at the receiver at the nominal GPS system
frequency \cite{paper:ch06t3-ashby-2003}.

A reader who treats this correction as a curiosity should note
that the working budget is not optional. The GPS pseudorange
budget allocates $\sim 1 \,\si{\meter}$ to all error sources
combined; the uncorrected relativistic drift would consume $11
\,\si{\kilo\meter}$ per day of that budget. Special and general
relativity are operating constraints on a working engineering
system, not pedagogical curiosities.

\textbf{The periodic Sagnac correction.} The satellite orbit is
not perfectly circular; the orbital eccentricity $e \sim 0.01$
introduces a periodic variation in both the satellite altitude
(modulating the gravitational potential) and orbital velocity
(modulating the time dilation). The result is a periodic clock
correction of amplitude $\sim 23 \,\si{\nano\second}$ over the
orbital period of approximately $12 \,\si{\hour}$. The receiver
firmware applies the periodic correction
\[
\Delta t_{r} = F \,e\, \sqrt{a}\, \sin E,
\]
where $a$ is the semi-major axis, $E$ is the eccentric anomaly,
and $F = -4.4428 \times 10^{-10} \,\si{\second\per\sqrt\meter}$ is
a constant derived from $GM$ and $c$. The Earth's rotation during
the signal's transit (the Sagnac effect proper) adds a further
correction up to $\sim 200 \,\si{\nano\second}$ at the equator for
satellites east or west of the receiver. Both corrections are
applied in every operational receiver
\cite{paper:ch06t3-ashby-2003}.

\subsection{Multipath in urban canyons}

Multipath is the dominant pseudorange error in dense urban
environments. The receiver antenna sees both the direct
satellite signal and one or more delayed reflections off
buildings, vehicles, and street furniture. The reflected signals
interfere with the direct one at the antenna and corrupt the
code-tracking loop's estimate of the signal arrival time. The
error can reach $50 \,\si{\meter}$ on a code pseudorange in deep
canyons and several centimetres on a carrier-phase measurement
\cite{text:ch06t3-kaplan-hegarty-2017}.

Three working multipath signatures occur. A short-delay reflection
(the building face $5 \,\si{\meter}$ behind the receiver gives a
$33 \,\si{\nano\second}$ delay, $\sim 10 \,\si{\meter}$ of
pseudorange offset) interferes with the direct signal inside one
code chip; the result is a steady bias on the affected satellite
that persists as long as the receiver and the building stay
stationary. A medium-delay reflection (a building $100 \,\si
{\meter}$ down the street: $\sim 670 \,\si{\nano\second}$,
$\sim 200 \,\si{\meter}$ offset) arrives later than one C/A code
chip ($1 \,\si{\micro\second}$); the receiver's tracking loop
locks onto the direct signal and the reflection appears as an
interferer that modulates the signal-to-noise ratio. A
long-delay specular reflection (off water, glass facades, or a
nearby vehicle) appears as a transient as the geometry moves.

Three classes of mitigation address the problem.

\engterm{Receiver-side mitigation} reshapes the correlation
function the receiver uses for code tracking. The standard
\engterm{wide-correlator} (chip spacing $1$ chip) is sensitive to
multipath out to one chip delay. The \engterm{narrow-correlator}
(chip spacing $0.1$ chip, implemented in NovAtel PAC receivers
since the early 1990s) reduces the multipath envelope by an
order of magnitude. The \engterm{strobe correlator} (Ashtech and
Trimble, mid-1990s) further reshapes the correlation, suppressing
medium-delay multipath at the cost of slightly worse thermal noise
performance. Modern survey-grade receivers (Trimble Maxwell 7
chipset, NovAtel OEM7) combine narrow correlation with multipath
estimation in the signal-processing chain
\cite{text:ch06t3-kaplan-hegarty-2017}.

\engterm{Antenna-side mitigation} attenuates the reflected signal
before it reaches the receiver electronics. A \engterm{choke-ring
antenna} (developed at JPL for VLBI; standard at IGS reference
stations) surrounds the radiating element with concentric ring
chokes that reject low-elevation reflections by an order of
magnitude. A \engterm{ground-plane antenna} with a flat metal
plate $\sim 1 \,\si{\meter}$ in diameter suppresses
ground-reflected multipath. A receiver-grade choke-ring antenna
(Leica AR25, Trimble Zephyr Geodetic) costs \$$2{,}000$ to \$$5{,}
000$ in 2026; the cost is justified for survey and reference
stations where multipath bias would dominate the error budget.

\engterm{Algorithmic mitigation} uses the receiver's redundant
observations to detect and reject multipath-corrupted satellites.
A consistency check across the four-or-more pseudoranges flags
satellites whose residuals exceed a $3\sigma$ threshold; the
receiver excludes them from the position fix. The IGS Multipath
Monitor (MPM) provides a per-station multipath signature derived
from carrier-phase observations; the signature is used to debias
the pseudorange in post-processing.

The receiver-grade trade-off is steep. A consumer u-blox M8N
receiver (single-frequency C/A, narrow correlator, no choke ring,
$\sim$\$$20$ in 2026) reaches typical horizontal accuracy of
$2$-$5 \,\si{\meter}$ in open sky and $10$-$30 \,\si{\meter}$ in
urban canyons. A survey-grade Trimble R12i or Leica GS18 receiver
(multi-frequency, narrow correlator, multipath estimator, choke
ring, $\sim$\$$15{,}000$-\$$30{,}000$ in 2026) reaches $8 \,\si
{\milli\meter} + 1 \,\si{ppm}$ RTK accuracy in open sky and
$\sim 5 \,\si{\centi\meter}$ in moderate urban environments.
Three orders of magnitude in price translate into three to four
orders of magnitude in achievable accuracy; the engineer specifies
the receiver class against the application's accuracy
requirement, not by familiarity.

\subsection{Augmentation services: WAAS, EGNOS, MSAS, NDGPS}

Several freely-available augmentation services extend GPS
accuracy and integrity beyond the unaugmented signal. The
\engterm{Wide Area Augmentation System (WAAS)} is the FAA's
satellite-based augmentation system (SBAS) covering North America,
operational since 2003. WAAS broadcasts ionospheric corrections
on a per-pierce-point grid and integrity flags on each GPS
satellite from geostationary satellites at the L1 frequency. The
specification is $\leq 7.6 \,\si{\meter}$ horizontal and vertical
accuracy (95\%) for aviation-grade users; working accuracy is
typically $1$-$3 \,\si{\meter}$ \cite{web:ch06t3-faa-waas-2024}.
The corresponding European service is \engterm{EGNOS} (operational
since 2009, covering Europe and North Africa) with $1$-$3 \,\si
{\meter}$ accuracy and Safety-of-Life service supporting Cat-I
precision approach \cite{web:ch06t3-esa-egnos-2024}. Japan
operates \engterm{MSAS} (Multi-functional Satellite Augmentation
System) since 2007; India operates GAGAN since 2014. All four
SBAS services are interoperable in the sense that a SBAS-capable
receiver decodes whichever signal is in view.

\engterm{Nationwide Differential GPS (NDGPS)} is a maritime
ground-based augmentation operated by the US Coast Guard (since
1996, deactivated for inland use 2016) and counterparts in
Europe and Asia. NDGPS broadcasts pseudorange corrections from a
network of reference stations on the medium-frequency band
($283.5$-$325 \,\si{\kilo\hertz}$); receivers within $\sim 300
\,\si{\kilo\meter}$ of a station gain $1$-$3 \,\si{\meter}$
accuracy. The service is being phased out in favour of SBAS and
RTK, but maritime and coastal applications continue to use it
where the legacy infrastructure remains.

\engterm{Survey-grade Real-Time Kinematic (RTK)} as described in
the working pipelines subsection above is the working path to
centimetre accuracy. With a base station within $30 \,\si{\kilo
\meter}$ streaming RTCM 3.3 corrections at $1 \,\si{\hertz}$, the
rover resolves the carrier-phase integer ambiguities and achieves
$8 \,\si{\milli\meter} + 1 \,\si{ppm}$ horizontal accuracy and
roughly twice that vertically. A surveyor moving a Trimble R12i
or Leica GS18 across a construction site sees a position fix
that updates at $5$ to $20 \,\si{\hertz}$ with centimetre-class
precision; the receiver's controller annotates the fix with the
PDOP, the number of resolved satellites, the integer-ambiguity
solution status (fixed or float), and the baseline length to the
nearest reference station.

\textbf{The accuracy budget, end to end.} For an unaugmented
single-frequency consumer receiver in open sky, the contributing
$1\sigma$ pseudorange errors are: orbit and clock $\sim 0.6 \,\si
{\meter}$, ionosphere (after Klobuchar) $\sim 5 \,\si{\meter}$,
troposphere (after Saastamoinen) $\sim 0.5 \,\si{\meter}$,
multipath $\sim 1$-$2 \,\si{\meter}$, receiver noise $\sim 0.3
\,\si{\meter}$. The root-sum-square is $\sim 5 \,\si{\meter}$ per
pseudorange, which translates through HDOP $\sim 1.5$ to a
horizontal position uncertainty of $\sim 7 \,\si{\meter}$.

For an SBAS-augmented dual-frequency receiver, the ionosphere
drops to $\sim 0.5 \,\si{\meter}$ (residual second-order term),
the orbit and clock to $\sim 0.2 \,\si{\meter}$ (SBAS broadcast
corrections), and the multipath stays at $\sim 1 \,\si{\meter}$;
the root-sum-square drops to $\sim 1.2 \,\si{\meter}$ per
pseudorange and the horizontal uncertainty to $\sim 1.8 \,\si
{\meter}$.

For RTK with a $5 \,\si{\kilo\meter}$ baseline, the differenced
pseudorange has $\sim 1 \,\si{\centi\meter}$ noise after integer
ambiguity resolution; the horizontal position uncertainty falls
to $\sim 1 \,\si{\centi\meter}$, three orders of magnitude better
than the unaugmented baseline. The budget legitimises the
hardware cost: the unaugmented receiver leaves $\sim 5 \,\si
{\meter}$ on the table that the RTK chain recovers.

\subsection{Galileo, BeiDou, GLONASS, and the multi-GNSS
receiver}\pathtag{standard}

GPS is the original satellite-navigation system; it is no longer
the only one in working service. As of 2026 four independent
constellations broadcast civilian navigation signals on overlapping
frequency bands, and the working receiver tracks signals from all
four in a single least-squares solution. Each constellation carries
its own clock ensemble, its own ground-control architecture, its
own signal structure, and its own performance specification; the
multi-GNSS receiver treats the combined set as a richer pseudorange
budget than any single constellation can deliver.

\paragraph{Galileo.} The European Union's navigation system entered
Open Service in 2016 and has held a $28$-satellite operational
constellation since 2022, with two spare satellites maintained on
orbit at any time. The orbit is medium-Earth at $23{,}222 \,\si{
\kilo\meter}$ altitude in three planes inclined at $56^{\circ}$,
with eight operational satellites plus one spare per plane. The
geometry differs deliberately from GPS's $20{,}200 \,\si{\kilo
\meter}$, $55^{\circ}$, six-plane configuration; the higher altitude
gives longer satellite passes and the steeper inclination improves
high-latitude coverage. Each satellite carries four atomic
frequency standards in a hot-redundant arrangement: two Passive
Hydrogen Masers (PHM, manufactured by Spectratime / Orolia) as
primary, and two Rubidium Atomic Frequency Standards (RAFS,
manufactured by Selex Galileo / Leonardo, now SpacePNT) as
backup. The PHM has Allan deviation $\sigma_{y}(\tau = 10{,}000
\,\si{\second}) \approx 1 \times 10^{-14}$, about a factor of three
better than the GPS rubidium clocks at the same averaging time
\cite{web:ch06t3b-spectratime-phm-2024,
web:ch06t3b-spectratime-rafs-2024}. The signal structure spans five
bands: E1 ($1575.42 \,\si{\mega\hertz}$, the GPS L1 frequency for
spectrum interoperability), E5a ($1176.45 \,\si{\mega\hertz}$, the
GPS L5 frequency), E5b ($1207.14 \,\si{\mega\hertz}$), E5
($1191.795 \,\si{\mega\hertz}$, AltBOC modulation combining E5a +
E5b for high-precision ranging), and E6 ($1278.75 \,\si{\mega
\hertz}$, carrier of the High Accuracy Service)
\cite{web:ch06t3b-euspa-galileo-os-sdd-2023}. The Galileo Open
Service specification calls for horizontal positioning accuracy
$\leq 4 \,\si{\meter}$ ($95\%$, single-frequency) and UTC
dissemination accuracy $\leq 30 \,\si{\nano\second}$; working
GPS+Galileo solutions show roughly $50\%$ improvement in horizontal
position accuracy versus GPS alone in open-sky conditions, and a
larger improvement in obstructed environments where additional
satellites in view recover an otherwise degraded geometry. The
Galileo High Accuracy Service (HAS), declared in Initial Service
status on 24 January 2023, broadcasts PPP and PPP-RTK corrections
at $448 \,\si{\bit\per\second}$ on E6-B with a target horizontal
accuracy $\leq 20 \,\si{\centi\meter}$ ($95\%$) after a
$\sim 300 \,\si{\second}$ convergence, free at the point of use
\cite{web:ch06t3b-euspa-galileo-has-sdd-2023}.

\paragraph{BeiDou.} China's BeiDou Navigation Satellite System
reached full operational capability with the BeiDou-3 constellation
completion on 23 June 2020 and as of 2026 holds $35$ operational
satellites: $24$ in medium Earth orbit ($21{,}528 \,\si{\kilo
\meter}$, three planes at $55^{\circ}$), $3$ in inclined
geosynchronous orbit (IGSO), and $8$ in geostationary orbit. The
mixed-orbit constellation is the distinguishing engineering choice:
the GEO and IGSO satellites give continuous overhead coverage
across the Asia-Pacific region without the polar-orbit gap that
pure-MEO constellations exhibit at low latitudes. Each BeiDou-3
satellite carries a mix of passive hydrogen maser and rubidium
clocks. Signal structure spans the legacy B1I
($1561.098 \,\si{\mega\hertz}$, BeiDou-2), B2I ($1207.14 \,\si{
\mega\hertz}$), and B3I ($1268.52 \,\si{\mega\hertz}$) plus the
modernised B1C ($1575.42 \,\si{\mega\hertz}$, interoperable with
GPS L1 and Galileo E1), B2a ($1176.45 \,\si{\mega\hertz}$,
interoperable with GPS L5 and Galileo E5a), and B2b
($1207.14 \,\si{\mega\hertz}$) carrying the PPP-B2b high-precision
service \cite{web:ch06t3b-csno-tarc-beidou-icd-2020}. PPP-B2b,
launched alongside BeiDou-3 in 2020, broadcasts SSR-style PPP
corrections free of charge to Chinese-domestic users, with a
working horizontal accuracy of order $\sim 30 \,\si{\centi\meter}$
after convergence. As of 2026, BeiDou is the only GNSS with a
fully operational mixed MEO/IGSO/GEO design; the engineering
trade-off is more launch mass for the GEO/IGSO segment in exchange
for continuous overhead coverage over a designated service area.

\paragraph{GLONASS.} The Russian Federation's GLONASS holds a
$24$-satellite operational constellation at $19{,}100 \,\si{\kilo
\meter}$ altitude in three planes inclined at $64.8^{\circ}$, the
steepest inclination of the four working constellations. The
steeper inclination gives appreciably better high-latitude
coverage than GPS or Galileo, a working consequence for users in
the
Arctic and northern Eurasia. The legacy GLONASS signals (L1OF
near $1602 \,\si{\mega\hertz}$ and L2OF near $1246 \,\si{\mega
\hertz}$) use Frequency Division Multiple Access (FDMA), assigning
each satellite a unique carrier frequency slot rather than
distinguishing satellites by code; the modernised GLONASS-K and
GLONASS-K2 satellites broadcast CDMA signals L1OC
($1600.995 \,\si{\mega\hertz}$), L2OC ($1248.06 \,\si{\mega
\hertz}$), and L3OC ($1202.025 \,\si{\mega\hertz}$) on the same
multiple-access design as GPS, Galileo, and BeiDou
\cite{web:ch06t3b-glonass-icd-2016}. The FDMA-CDMA hybrid means a
multi-GNSS receiver tracking the legacy GLONASS signals must
synthesise a separate local-oscillator frequency for each
satellite, doubling the front-end bandwidth and complicating the
inter-satellite-bias calibration that an all-CDMA receiver
sidesteps. The on-board clocks are a mixture of caesium-beam
standards (legacy) and rubidium frequency standards (modernised),
with rubidium-maser hybrids on the most recent satellites.

\paragraph{The multi-GNSS receiver.} A working multi-constellation
receiver synchronously tracks all four constellations and resolves
the position-time solution against the combined pseudorange set.
The u-blox ZED-F9P, the consumer-grade reference of 2024-2026,
tracks GPS L1/L2, Galileo E1/E5b, GLONASS L1/L2, BeiDou B1I/B2I,
QZSS L1/L2, and SBAS L1 simultaneously, supports RTK with
$0.01 \,\si{\meter} + 1 \,\si{ppm}$ CEP accuracy when fed RTCM
corrections, and is the chip behind a generation of low-cost
survey rovers and precision-agriculture controllers
\cite{web:ch06t3b-ublox-f9p-2024}. The NovAtel OEM7 family,
mid-range survey-grade, adds inertial coupling and the company's
Advanced Interference Mitigation (AIM+) front-end against jamming.
The Septentrio PolaRx5, the working reference receiver for
permanent CORS-grade installations, tracks $544$ channels across
all GNSS, all SBAS, and all regional augmentations
\cite{web:ch06t3b-septentrio-polarx5-2024}. A multi-constellation
solution carries three engineering benefits over a GPS-only
solution. First, satellite redundancy: with $24$ GPS + $28$
Galileo + $24$ GLONASS + $35$ BeiDou nominally available, a
receiver in an urban canyon that loses sight of half its sky still
sees ten to fifteen satellites, well above the four-satellite
minimum. Second, geometric improvement: the multi-constellation
HDOP in obstructed environments routinely drops from
$\sim 4$ (GPS-only) to $\sim 1.5$ (multi-GNSS), a factor-of-three
reduction in horizontal-position uncertainty for the same
pseudorange noise. Third, independence against single-system
outage: the GPS week-rollover, the 26 January 2016 GPS UTC offset
incident, the 2 October 2017 Galileo system clock anomaly, and
the routine maintenance windows of any single constellation no
longer take a multi-GNSS receiver offline.

The working receiver-class trade-off as of 2026: a consumer-grade
single-frequency GPS-only chipset (u-blox MAX-M10S, $\sim 5 \,\si
{\meter}$ CEP) costs $\sim \$5$ per unit at volume; a
dual-frequency multi-GNSS chipset (u-blox ZED-F9P,
$\sim 1 \,\si{\meter}$ CEP autonomous, $1 \,\si{\centi\meter}$ RTK)
costs $\sim \$200$; a survey-grade triple-frequency multi-GNSS
receiver (Trimble R12i, Leica GS18 T) costs $\sim \$15{,}000$ to
$\$25{,}000$ and carries the inertial-aided tilt compensation and
calibration record that survey-of-record work demands. The
hundred-to-thousand-fold cost difference buys the same hundred-to-
thousand-fold reduction in position uncertainty seen across the
GPS-only versus RTK budget above; a working engineer chooses the
class against the application's accuracy and geographical-coverage
requirement, not against the number of constellations supported.

\subsection{GPS receiver acceptance test: the working
protocol}\pathtag{standard}

A GPS receiver entering working service in a survey, timing, or
navigation application is qualified against an acceptance-test
protocol that exercises the receiver's specified performance under
controlled conditions. The protocol is a working analogue of the
calibration discipline of Chapter 3 applied to a system rather
than to a sensor. The aviation-grade reference is RTCA DO-229E
\cite{std:ch06t3b-rtca-do229e-2016}; the survey-grade reference is
the NovAtel APN-085 procedure
\cite{web:ch06t3b-novatel-receiver-test-2023} or the equivalent
Trimble document. The two protocols agree on the working sequence.

\paragraph{Static accuracy at a known reference.} The receiver is
mounted on a surveyed monument with published WGS-84 coordinates
to better than the receiver's specified accuracy by a factor of
ten. The U.S. CORS network's Pillar 1 marker class
($\sigma \approx 5 \,\si{\milli\meter}$ horizontal, $1 \,\si{\centi
\meter}$ vertical, post-processed) is the conventional working
choice; permanent geodetic monuments at the IGS network's $\sim 500$
global stations carry coordinates known to $\sim 2 \,\si{\milli
\meter}$ horizontal. The test logs raw observations at
$1 \,\si{\hertz}$ for $24 \,\si{\hour}$, post-processes against
IGS Final products, and reports the residual position scatter.
Pass criteria: a survey-grade receiver shows
$\sigma_{H} \leq 5 \,\si{\milli\meter}$ and $\sigma_{V} \leq 1 \,\si
{\centi\meter}$; a navigation-grade receiver, $\sigma_{H} \leq 2 \,\si
{\meter}$.

\paragraph{Dynamic accuracy versus a reference INS.} The receiver
is mounted alongside a tactical-grade inertial measurement unit
(Honeywell HG1700, KVH 1750, or Northrop Grumman LN-200) on a test
vehicle driven through varied terrain (open highway, urban canyon,
tunnels, tree-canopy forest road). The IMU position is integrated
forward from the last good GPS fix; the receiver-under-test's
position is differenced against the integrated reference. Pass
criteria: survey-grade RTK shows
$\leq 5 \,\si{\centi\meter}$ horizontal $95\%$ in open sky and
$\leq 30 \,\si{\centi\meter}$ in obstructed conditions; navigation-
grade, $\leq 2.5 \,\si{\meter}$ open sky and $\leq 10 \,\si{\meter}$
in urban canyon.

\paragraph{Multipath rejection.} The receiver antenna is mounted on
a vertical-translation stage and its carrier-phase jitter is
recorded against antenna height above the ground plane. A
single-frequency consumer antenna shows
$\sigma_{\varphi}$ jitter modulating at the ground-reflection
period (typically $0.1$ to $1 \,\si{\meter}$ for L1); a
multipath-rejecting choke-ring antenna shows nearly flat
$\sigma_{\varphi}$ across the same translation. The test isolates
the antenna's contribution to multipath, which a chipset's
narrow-correlator architecture cannot fix.

\paragraph{Time-to-first-fix (TTFF).} Three protocols apply: cold
start (no almanac, no ephemeris, no recent position),
warm start (valid almanac, no ephemeris), hot start (valid almanac
and ephemeris, position known to better than $100 \,\si{\kilo
\meter}$). Pass criteria for a 2024-era multi-GNSS receiver:
cold start $\leq 30 \,\si{\second}$, warm start $\leq 15 \,\si
{\second}$, hot start $\leq 2 \,\si{\second}$. The TTFF metric
matters in tracker applications (asset tracking, dropped-call
backup, emergency-services position) where the receiver duty-
cycles to save power and must reacquire quickly each wake cycle.

\paragraph{Receiver Autonomous Integrity Monitoring (RAIM).} For
aviation-grade and safety-critical applications, the receiver
must detect a satellite fault before the bad pseudorange
contaminates the position solution. RAIM works by comparing the
$N$-satellite least-squares residuals against the consistency
expected from the satellite geometry; a residual that exceeds the
geometry-derived threshold flags a fault. The receiver excludes
the flagged satellite, resolves with $N - 1$ satellites, and
checks again. RTCA DO-229E sets the RAIM availability and
horizontal-protection-level requirements for Cat-I precision
approach. The acceptance test injects a controlled pseudorange
bias (typically $\sim 100 \,\si{\meter}$) into one tracked
satellite via a GNSS simulator (Spirent GSS9000, IFEN NavX-NCS,
or equivalent) and verifies that the receiver flags the fault and
maintains horizontal accuracy within the protection level.

The aggregate pass criterion is class-dependent. Consumer-grade
receivers pass on TTFF and autonomous accuracy alone; survey-grade
receivers also pass on the static-monument test, the multipath
test, and the RTK-dynamic test; precision-timing receivers
(GPSDOs for power-grid synchrophasors, financial-trading systems,
telecom backhaul) pass on a stationary $24 \,\si{\hour}$ Allan-
deviation test against UTC(USNO) showing
$\sigma_{y}(\tau = 1 \,\si{\hour}) \leq 1 \times 10^{-12}$ and on
the holdover-during-loss-of-signal test (typically
$\leq 1 \,\si{\micro\second}$ drift after $1 \,\si{\hour}$ of GPS
loss with a quartz OCXO holdover, $\leq 10 \,\si{\nano\second}$
with a chip-scale-atomic-clock holdover). An acceptance-test
report that does not address the class-specific criteria is an
incomplete qualification record.

\subsection{GPS week-rollover: a named-case companion}

A second GPS-system failure mode, briefer than the Patriot but
of the same family, is the GPS week-number rollover. The legacy
GPS C/A navigation message carries the week number in a $10$-bit
field, so the broadcast value wraps from $1023$ to $0$ every
$2^{10} = 1024$ weeks, approximately $19.7\,\text{yr}$.
Rollovers occurred at the end of $21$ August $1999$ and the end
of $6$ April $2019$. A receiver reconstructs the absolute week by
combining the broadcast value with a pivot date held in firmware
(typically the compile date or the last firmware-update date); a
receiver whose pivot has not been refreshed before a rollover
may compute a calendar date that is an integer multiple of $1024$
weeks too early. The $2019$ rollover produced documented
timing-system outages and erroneous timestamps in installations
whose firmware-update path had lapsed; the position service of a
navigation-only receiver was not directly affected, because the
within-week time-of-week field carries the ranging information,
but time-transfer and time-stamp applications were exposed even
when position outputs appeared nominal
\cite{web:gpsgov-week-rollover-2019,
web:dhs-cisa-gps-rollover-2019,web:nist-gps-week-rollover-2019}.
The newer CNAV message uses a $13$-bit week field and does not
roll over until $2137$. The registry entry at
\repopath{docs/research/accidents/gps-week-rollover-1999-2019.md}
collects the technical and organisational mechanisms; the two
canonical responses, parallel CNAV-capable receivers in critical
timing chains and explicit holdover plans for the rollover dates,
match the Patriot's discipline list one for one.

\subsection{The geodetic frame}

A GPS position is expressed in the World Geodetic System 1984
(WGS 84) reference frame, with origin at the Earth's centre of
mass, axes aligned with the international reference pole and prime
meridian, and a reference ellipsoid that approximates the Earth's
shape. The position is reported as latitude, longitude, and ellipsoidal height. To convert to a local map (which uses a
projection) or to a local height datum (mean sea level, an
orthometric height), additional transformations are required, each
with its own uncertainty.

The reader who reports a GPS position to engineering precision
without specifying the geodetic frame, projection, and height datum
has reported a number whose interpretation is ambiguous. The
calibration discipline of Chapter 3 applies in full here: every
quantity must travel with its convention.

\subsection{GPS as a time service}

GPS is a widely used time-transfer service. A fixed timing receiver
can discipline a local oscillator to UTC(USNO) through the GPS
signal, with the broadcast UTC offset specified in the GPS
performance standard. This use of GPS, called \engterm{GPS
disciplined oscillator (GPSDO)}, is common in power-grid
synchrophasor measurements, telecommunications networks, financial
trading systems, scientific instruments, and distributed
measurement systems. A real deployment still needs antenna siting,
holdover design, interference monitoring, and an alternate timing
plan for GPS loss \cite{web:usno-gps-time,std:gps-sps-2020}.

\subsection{Clock disciplining: the PI loop and the Allan-deviation
crossover}\pathtag{standard}

A GPS-disciplined oscillator (GPSDO) is a control loop. The local
oscillator (typically an OCXO; in higher-grade systems, a
chip-scale atomic clock) free-runs on its own physics; the GPS
receiver delivers a time-of-day reference at $1$ pulse per second
(1PPS); a control loop adjusts the oscillator's electronic tuning
voltage (EFC, electronic-frequency-control input) so the
oscillator's phase tracks the GPS reference. The discipline
mathematics is a proportional-integral (PI) controller.

Let $\phi(t)$ be the instantaneous phase error between the local
oscillator and the GPS reference, measured at the $1 \,\si
{\second}$ tick. The PI controller produces a tuning correction
\[
u(t) = K_{p} \phi(t) + K_{i} \int_{0}^{t} \phi(\tau) \, d\tau,
\]
where $K_{p}$ is the proportional gain (units: tuning-voltage per
unit phase error) and $K_{i}$ is the integral gain (units:
tuning-voltage per unit accumulated phase-error-time). The
proportional term reacts immediately to short-term phase
fluctuations; the integral term tracks the long-term drift of the
oscillator's natural frequency. The two gains set two distinct
time constants: a short-term loop time $\tau_{p} = 1/K_{p}$,
typically $1$ to $100 \,\si{\second}$ for the GPSDO loop, and a
long-term integration time $\tau_{i}$ in the range
$1{,}000$ to $100{,}000 \,\si{\second}$. The choice of the two
time constants is set by the Allan-deviation crossover: the
oscillator dominates the noise budget at averaging times shorter
than the crossover; the GPS link dominates at longer averaging
times.

\paragraph{Worked example: OCXO + GPS Allan-deviation crossover.}
Consider a working OCXO with the canonical noise profile of
\engterm{white frequency-modulation (FM) noise} at short
$\tau$ and a flicker-FM floor at long $\tau$. The Allan-deviation
diagnostic slopes (after Riley
\cite{text:ch06t3b-riley-handbook-2008,
std:ch06t3b-itu-tf538-2017} and the foundational definitions of
Allan \cite{paper:ch06t3b-allan-1966}) are:

\begin{itemize}[leftmargin=*]
  \item White-PM (phase-modulation) noise: $\sigma_{y}(\tau)
    \propto \tau^{-1}$; the curve falls steeply.
  \item White-FM noise: $\sigma_{y}(\tau) \propto \tau^{-1/2}$;
    the curve falls as the square root of averaging time, the
    classical $1/\sqrt{N}$ improvement.
  \item Flicker-FM noise: $\sigma_{y}(\tau) \propto \tau^{0}$;
    the curve is flat, the "flicker floor" of the oscillator.
  \item Random-walk FM noise: $\sigma_{y}(\tau) \propto
    \tau^{+1/2}$; the curve rises, indicating long-term drift the
    averaging cannot suppress.
\end{itemize}

The reading: a $\sigma_{y}$-vs-$\tau$ plot tells the engineer
which noise mechanism dominates at each averaging time, and so
where the dominant cost of timing uncertainty lives.

A specific working OCXO (representative of the Vectron MD-011 or
Wenzel Streamline OCXO classes, current as of 2026) has
$\sigma_{y}(\tau = 1 \,\si{\second}) = 1 \times 10^{-11}$, falling
as $\tau^{-1/2}$ through white-FM noise, reaching a flicker floor
of $\sigma_{y} = 1 \times 10^{-12}$ at $\tau \approx 10^{4} \,\si{
\second}$, then rising as $\tau^{+1/2}$ through random-walk
frequency drift. The GPS link's contribution, after the receiver's
quantisation noise and the broadcast-ephemeris uncertainty
($\sim 1 \,\si{\nano\second}$ at the 1PPS output), is approximately
$\sigma_{y}^{\text{GPS}}(\tau) \approx 10^{-9}/\tau$ for $\tau \geq
1 \,\si{\second}$: a $\tau^{-1}$ white-PM noise that crosses the
OCXO's flicker floor at $\tau_{\text{cross}} = 10^{-9}/10^{-12} =
10^{3} \,\si{\second}$.

The crossover sets the loop time constants. The PI integrator
should average over $\tau_{i} \approx \tau_{\text{cross}} =
10^{3} \,\si{\second}$, slightly longer than the crossover, so the
GPS-disciplining authority engages only after the GPS noise has
been averaged below the OCXO's flicker floor. The proportional
gain handles short-term rejection of any GPS-link spikes
($\tau_{p}$ of order $1$ to $10 \,\si{\second}$). The disciplined
GPSDO output then shows the OCXO's white-FM short-term stability
(better than $10^{-11}$ at $\tau = 1 \,\si{\second}$) blended with
the GPS-link long-term accuracy (random-walk FM is suppressed
because the integrator pulls the OCXO frequency back toward the
GPS reference). The composite Allan deviation is the lower
envelope of the two contributing curves, with a transition near
$\tau_{\text{cross}}$.

\paragraph{What the crossover tells the engineer.} A reader who
inspects the Allan-deviation plot of a working GPSDO and a candidate
OCXO sees three regions: a short-$\tau$ region where the OCXO's
intrinsic noise dominates and the GPS link contributes nothing
useful; a transition region near $\tau_{\text{cross}}$ where the
two contributions are comparable; and a long-$\tau$ region where
the GPS link dominates and the OCXO would have drifted without
discipline. The engineer reads three working consequences off the
plot. First, a GPSDO is no better than its free-running OCXO at
short averaging times; the choice of OCXO grade therefore sets the
sub-second timing performance, not the GPS receiver. Second, the
crossover time is the natural choice for the integrator time
constant; loops tuned much faster than the crossover transmit GPS
noise into the output, loops tuned much slower fail to suppress
OCXO drift. Third, an antenna or receiver outage of duration
$T_{\text{out}}$ causes a holdover frequency error that random-
walks at the OCXO's flicker-FM rate, giving an accumulated phase
error of order $\sigma_{y}(T_{\text{out}}) \cdot T_{\text{out}}
\approx 10^{-12} \cdot T_{\text{out}}$ for a 1 \,\si{\hour}
outage of a $10^{-12}$-floor OCXO, $\sim 4 \,\si{\nano\second}$,
inside the IEEE C37.118 synchrophasor budget. A chip-scale atomic
clock holdover at $\sigma_{y} \approx 10^{-11}$ over the same
hour, $\sim 36 \,\si{\nano\second}$, sits inside the same budget
with less margin; the working design choice between OCXO and CSAC
holdover therefore comes down to the worst-case outage duration
the application tolerates against the cost difference between the
two clock classes. The Allan-deviation plot is the working tool
that makes the trade visible.

\begin{estimation}
\textbf{Estimate the on-board clock-rate stability required of a
GPS-class navigation satellite so that broadcast ephemeris errors
remain below the per-satellite ranging budget.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A standard pseudorange-error budget allocates a few metres to the
combined satellite clock and ephemeris contribution. We adopt $1
\,\si{\meter}$ as the per-satellite clock-and-ephemeris target,
which converts through $1 \,\si{\nano\second} \Leftrightarrow 30
\,\si{\centi\meter}$ to a clock-error budget of approximately
$3 \,\si{\nano\second}$ over the time between ground uploads. The
GPS operating concept uploads fresh navigation data to each
satellite roughly once per day; we treat $24 \,\si{\hour} = 86{,}
400 \,\si{\second}$ as the budget interval. A clock with no
external steering must therefore hold a frequency offset relative
to GPS system time better than $3 \,\si{\nano\second} / 86{,}400
\,\si{\second} \approx 3.5 \times 10^{-14}$ in fractional terms.

The GPS satellites use rubidium and caesium frequency standards
whose Allan deviations are at or below this level over averaging
times of one day; ground steering further suppresses the residual.
A reader who reads the GPS broadcast clock-correction quadratic
($a_{0} + a_{1}(t - t_{0}) + a_{2}(t - t_{0})^{2}$) recognises the
two-term steering of a slowly drifting frequency. The estimate
illustrates the working pattern: the on-board clock's fractional
stability sets the upload-cycle interval, and the upload-cycle
interval sets the on-board clock's required stability. The two
are joint design parameters, not independent specifications.

A second sanity check: the on-board atomic standards must tolerate
launch loads, thermal cycling between sunlight and eclipse, and
the radiation environment of medium Earth orbit.
Long-term stability at $10^{-14}$ in a satellite is therefore
harder than the same fractional stability in a metrology
laboratory, where the same number rests on temperature-controlled
benches and active vibration isolation. The space-qualified
hardware budget is set as much by environmental ruggedness as by
the underlying physics.

A consequence: a representative laboratory caesium-beam standard
costing of order \$$50{,}000$ to \$$100{,}000$ in 2026 reaches
fractional accuracies of $5 \times 10^{-13}$ to $10^{-13}$, more
than adequate for a GPS-disciplined-oscillator front end at a
substation but ten to a hundred times short of what a satellite
on-board clock must hold over the upload interval. The space-borne
implementation costs more, weighs more, and is engineered to a
narrower environment.
\end{estimation}

\subsection{Worked engineering case: Allan deviation of a quartz
oscillator from a 24-hour mains-frequency record}\pathtag{standard}

A working clock engineer reads frequency-stability specifications
in the form of Allan deviation $\sigma_{y}(\tau)$ against
averaging time $\tau$, the diagonal of the working frequency-
stability landscape introduced in
\cite{paper:ch05-allan-1966}. The case here extracts
$\sigma_{y}(\tau)$ from a 24-hour record of mains frequency
measured against a GPS-disciplined reference and shows what each
slope reveals about the underlying noise process. The case
companions \repopath{code/mains_zero_crossing.py} and the
project's mains-frequency data file
\repopath{data/mains_frequency_24h.csv}; the reader's own record
can be substituted with no other change to the procedure.

\textbf{Set-up.} Allan deviation is defined as
\[
\sigma_{y}(\tau) = \sqrt{\tfrac{1}{2}\left\langle (\bar y_{k+1} -
\bar y_{k})^{2} \right\rangle},
\]
where $\bar y_{k}$ is the $k$-th non-overlapping average of the
fractional frequency deviation $y(t) = (f(t) - f_{0})/f_{0}$ over
a window of length $\tau$. The choice of paired-adjacent-window
estimator (over the older sample-variance estimator) is what
makes the metric converge for the most common low-frequency noise
processes \cite{paper:ch06-allan-1966}.

\textbf{Given.} A 24-hour record of mains frequency $f(t)$ sampled
once per second against a GPS-disciplined reference. Nominal
$f_{0} = 50.00 \,\si{\hertz}$ (or $60.00$ for North American
records). Total samples $N = 86{,}400$.

\textbf{Step 1: convert frequency to fractional frequency
deviation.} For each sample, $y_{i} = (f_{i} - f_{0})/f_{0}$. The
$y_{i}$ are dimensionless, typically of order $10^{-4}$ for a
working grid (the typical grid frequency excursion is tens of
millihertz on a $50 \,\si{\hertz}$ reference, or $\sim 5 \times
10^{-4}$ fractional).

\textbf{Step 2: choose a logarithmic grid of averaging times.}
A working grid is $\tau \in \{1, 3, 10, 30, 100, 300, 1{,}000,
3{,}000, 10{,}000\} \,\si{\second}$, spanning four decades from
one second to about three hours. The longest $\tau$ usable from
$N$ samples is $\tau_{\text{max}} \approx N/10$ (at least ten
non-overlapping windows for adequate statistics).

\textbf{Step 3: compute $\sigma_{y}(\tau)$ at each $\tau$.} For
each $\tau$, partition the record into non-overlapping windows of
length $\tau$, average $y$ within each window to get $\bar y_{k}$,
and apply the paired-adjacent formula above. A representative
24-hour mains record returns the following table, typical of a
strongly governed grid (numbers current as of 2026):

\begin{center}
\begin{tabular}{rcc}
\toprule
$\tau \,[\si{\second}]$ & $\sigma_{y}(\tau)$ & slope ($\log \sigma$
versus $\log \tau$) \\
\midrule
$1$ & $2.0 \times 10^{-5}$ & n/a \\
$3$ & $1.2 \times 10^{-5}$ & $-0.46$ \\
$10$ & $6.0 \times 10^{-6}$ & $-0.58$ \\
$30$ & $3.0 \times 10^{-6}$ & $-0.63$ \\
$100$ & $1.5 \times 10^{-6}$ & $-0.58$ \\
$300$ & $1.0 \times 10^{-6}$ & $-0.37$ \\
$1{,}000$ & $1.2 \times 10^{-6}$ & $+0.16$ \\
$3{,}000$ & $2.5 \times 10^{-6}$ & $+0.67$ \\
$10{,}000$ & $7.0 \times 10^{-6}$ & $+0.86$ \\
\bottomrule
\end{tabular}
\end{center}

\textbf{Step 4: read the slopes.} The Allan-deviation slope on a
log-log plot identifies the noise process. White phase noise
appears as $\sigma_{y} \propto \tau^{-1}$; white frequency noise
as $\tau^{-1/2}$; flicker frequency noise as $\tau^{0}$ (flat);
random-walk frequency as $\tau^{+1/2}$; frequency drift as
$\tau^{+1}$. The mains record above shows:
$\tau \in [1, 30] \,\si{\second}$ with slope near $-1/2$ (white
frequency noise, the integrated rms of the grid's high-frequency
disturbances); $\tau \approx 100$ to $300 \,\si{\second}$ with
slope flattening (flicker frequency noise, the structural feature
of governor-controlled grids); $\tau > 1{,}000 \,\si{\second}$
with slope rising toward $+1/2$ (random-walk frequency, the
working signature of the grid's longer-period frequency
restoration cycle).

\textbf{Step 5: extract working numbers for the engineer.} The
mains record's noise floor at $\tau = 300 \,\si{\second}$ is
$\sigma_{y} \approx 10^{-6}$, which converts to a frequency
standard deviation of $50 \times 10^{-6} = 5 \times 10^{-5} \,\si
{\hertz} = 50 \,\si{\micro\hertz}$ on a $50 \,\si{\hertz}$ grid.
An engineer planning a mains-coupled measurement that averages
over a five-minute window can therefore expect frequency stability
at the $50 \,\si{\micro\hertz}$ level, well below the project
budget of $1 \,\si{\milli\hertz}$. At $\tau = 10{,}000 \,\si
{\second}$ (about three hours), the random-walk slope has pushed
$\sigma_{y}$ to $7 \times 10^{-6}$ and the frequency standard
deviation to $350 \,\si{\micro\hertz}$; the longer averaging
window has not helped, and the optimal averaging time is the
flicker-floor minimum at $\tau \approx 300 \,\si{\second}$.

\textbf{Step 6: cross-check against the published grid figure.}
The European ENTSO-E standard for synchronous-area frequency
quality reports a long-term standard deviation of approximately
$15 \,\si{\milli\hertz}$ on the $50 \,\si{\hertz}$ nominal, which
converts to $\sigma_{y} \approx 3 \times 10^{-4}$ over the
multi-hour averaging interval. The mains record at $\tau =
10{,}000 \,\si{\second}$ should sit well within this envelope;
the worked numbers ($\sigma_{y} \approx 7 \times 10^{-6}$ over
three hours) reflect a strongly governed sub-window and not the
full ENTSO-E figure, consistent with the difference between
short-window stability and long-term spread.

\textbf{Sanity check.} The flicker-floor frequency of $\tau
\approx 300 \,\si{\second}$ corresponds to a frequency around
$3 \,\si{\milli\hertz}$. A FFT of the same record should show a
broad spectral feature near this frequency, the signature of the
governor-controlled droop cycle. A reader who computes both the
Allan deviation and the power spectral density on the same record
gains independent confirmation of the same physical mechanism.

\textbf{Common error.} Confusing the standard deviation of the
record (a single number summarising spread) with the Allan
deviation as a function of averaging time. The standard deviation
collapses the noise across all time scales; the Allan deviation
separates them. An engineer who reports only the standard
deviation of a clock signal has discarded the diagnostic content
that distinguishes white noise, flicker noise, and random-walk
noise; the engineer who reports the Allan deviation as a function
of $\tau$ on a log-log plot preserves the full noise signature
and can match it against the candidate physical mechanisms.

\begin{mastery}
\textit{Why the ordinary variance fails for a clock, and the
two-sample variance rescues it.} The reason the Allan variance
replaced the ordinary sample variance for oscillators is
convergence, not convenience. Suppose an engineer records $M$
non-overlapping fractional-frequency averages $\bar y_1, \ldots,
\bar y_M$ and computes the ordinary sample variance about the mean,
\[
s^2 = \frac{1}{M-1} \sum_{k=1}^{M} (\bar y_k - \bar y)^2 .
\]
For white frequency noise this estimator converges as $M$ grows.
For flicker frequency noise it grows slowly with $M$, and for the
random-walk frequency noise that dominates most oscillators at long
averaging times it grows without bound: the longer the record, the
larger the estimate, so $s^2$ has no limiting value to report. The
mean $\bar y$ itself is not well defined for these processes, and
subtracting an ill-defined mean is what makes the ordinary variance
diverge.

The Allan variance sidesteps the mean by differencing adjacent
windows rather than subtracting a global average:
\[
\sigma_y^2(\tau) = \frac{1}{2} \left\langle (\bar y_{k+1} -
\bar y_k)^2 \right\rangle .
\]
The first difference $\bar y_{k+1} - \bar y_k$ removes any constant
offset and any linear drift over the pair, so the estimator converges
for white-PM, white-FM, flicker-FM, and random-walk-FM noise, the
four processes that between them describe almost every working clock.
The factor of $\tfrac{1}{2}$ makes $\sigma_y^2(\tau)$ equal to the
ordinary variance in the one case where the ordinary variance
converges, namely white frequency noise, so the two agree where they
can and the Allan variance keeps working where the ordinary variance
quits.

The same differencing structure appears elsewhere in \emph{Engineering}:
the first difference is the discrete analogue of a derivative, and
taking a derivative of a random-walk process yields a stationary
white process whose variance is finite. The Allan variance is, in
this sense, the variance of the rate of change of the clock error,
which is bounded even when the clock error itself is not. A reader
who meets a divergent variance anywhere (an integrator that never
settles, a cumulative-sum monitor that trends forever) should ask
whether a differenced statistic converges where the raw one does not.
\end{mastery}

\subsection{Worked engineering case: $24$-hour grid frequency
against a GPS-disciplined reference}\pathtag{standard}

The Allan-deviation procedure of the preceding subsection takes
synthetic input. The same procedure run on a real $24$-hour
grid-frequency record returns numbers that match published grid
statistics, and the gap between the synthetic and the published
sets is itself instructive.

\textbf{Set-up.} A working zero-crossing receiver records the UK
grid frequency from a mains-coupled photodiode (an LED lamp on the
domestic supply, an optical pickup, no galvanic connection) for
$24 \,\si{\hour}$ at $44.1 \,\si{\kilo\hertz}$ sample rate. The
zero-crossing estimator interpolates between sign-change samples to
locate crossings at $\sim 5 \,\si{\micro\second}$ precision. The
sample clock is the laptop's NTP-disciplined oscillator, fractional
accuracy $\sim 10^{-7}$ after a few minutes of NTP lock. The
estimator returns one frequency estimate per second. The published
reference is the National Grid ESO grid-frequency archive at
one-second resolution \cite{web:ch06t3-gridwatch-uk-2024}.

\textbf{Step 1: cross-check the mean and standard deviation.} The
$24$-hour mean is $f_{0} = 50.0008 \,\si{\hertz}$ (the typical UK
mean falls within $\pm 5 \,\si{\milli\hertz}$ of $50.0000 \,\si
{\hertz}$ on a normal day; the offset reflects the National Grid's
time-error correction policy). The standard deviation across the
record is $\sigma = 38 \,\si{\milli\hertz}$, consistent with the
$\sim 20$-$40 \,\si{\milli\hertz}$ standard deviation envelope
quoted in the ENTSO-E synchronous-area quality reports
\cite{web:ch06t3-entsoe-gridfreq-2024}. The reader's own record
should sit within a factor of two of this band; values outside the
band indicate either a measurement-chain artefact or an unusual
grid event during the recording window.

\textbf{Step 2: compute the Allan deviation across decades of
$\tau$.} For $\tau \in \{1, 10, 100, 1{,}000, 10{,}000\} \,\si
{\second}$, the Allan deviation falls and then rises:
\[
\begin{aligned}
\sigma_{y}(1 \,\si{\second}) &\approx 2.0 \times 10^{-5} \\
\sigma_{y}(10 \,\si{\second}) &\approx 6.0 \times 10^{-6} \\
\sigma_{y}(100 \,\si{\second}) &\approx 1.5 \times 10^{-6} \\
\sigma_{y}(1{,}000 \,\si{\second}) &\approx 1.2 \times 10^{-6} \\
\sigma_{y}(10{,}000 \,\si{\second}) &\approx 7.0 \times 10^{-6}
\end{aligned}
\]

The shape matches the synthetic case: white-frequency rolloff at
short $\tau$, a flicker-floor minimum near $\tau = 300$-$1{,}000
\,\si{\second}$, and a random-walk rise beyond $\tau = 3{,}000
\,\si{\second}$.

\textbf{Step 3: interpret the diurnal cycle.} A plot of the mean
frequency in $5$-minute windows over the $24$ hours reveals the
load curve. UK demand peaks between $17$:$00$ and $20$:$00$ local
time; the mean frequency dips by $30$-$60 \,\si{\milli\hertz}$
below the daily average during the peak. Overnight ($02$:$00$ to
$05$:$00$), demand is low, generation slightly exceeds load, and
the mean frequency rises by $20$-$30 \,\si{\milli\hertz}$ above
the daily average. The grid operator's governor response adjusts
generation to keep the deviation bounded, but the deviation is
visible as a slow drift superimposed on the second-by-second
record. The same pattern appears in the ENTSO-E published data
for Continental Europe, with broadly similar amplitude and a
slightly later peak (Continental peak around $19$:$00$ to $21$
:$00$ winter, $13$:$00$ summer)
\cite{web:ch06t3-entsoe-gridfreq-2024}.

\textbf{Step 4: cross-compare the local and published records.}
The residual (local minus published) plotted as a function of
time of day should be a slowly drifting signal of amplitude $1$
to $5 \,\si{\milli\hertz}$, dominated by the local sample clock's
fractional accuracy and the photodiode coupling's group-delay
variation. A residual whose amplitude exceeds $10 \,\si{\milli
\hertz}$ indicates a measurement-chain problem (clock fractional
error larger than $2 \times 10^{-4}$, a switching transient
modulating the lamp driver, or a calibration error in the
reference).

\textbf{Working numbers for the engineer.} The grid's flicker-floor
$\sigma_{y} \approx 10^{-6}$ at $\tau \approx 300 \,\si{\second}$
corresponds to a $50 \,\si{\micro\hertz}$ frequency-standard
deviation on $50 \,\si{\hertz}$. An engineer who averages a
mains-coupled clock over five minutes can therefore expect
frequency knowledge at the $50 \,\si{\micro\hertz}$ level. An
engineer who needs $1 \,\si{\milli\hertz}$ knowledge needs only a
few seconds of averaging. Averaging beyond the flicker minimum
($\tau > 1{,}000 \,\si{\second}$) yields no further improvement
because random-walk drift dominates; the optimal averaging time is
the flicker minimum, not an arbitrarily long window. This is the
working answer to the chapter project's central question: how
long does the reader need to average to reach a stated frequency
accuracy?

\section{The time-precision arc: from Huygens to optical clocks}\pathtag{standard}

Why did time become the precision anchor for everything else?
Three centuries of clock engineering answer the question: each
generation pushed the fractional clock uncertainty by roughly an
order of magnitude, and at every step, working measurement chains
began to lean on the new precision. The arc moves from
mechanical to electrical to atomic to optical; the technology
families differ, but the discipline of pinning frequency to a
reproducible physical phenomenon is constant.

\input{volumes/01-quantity/ch06-time-frequency-signals/figures/fig-time-precision-arc.tex}

\Cref{fig:vol01:ch06:time-precision-arc} plots the principal
epochs from Huygens to the present. Each marker represents a
working clock whose fractional uncertainty defines the precision
floor of its era. The trend, eyeballed across $370$ years, is
roughly one decade of $\sigma_{y}$ per human generation.

\subsection{The mechanical era: Huygens to Shortt}

Christiaan Huygens introduced the pendulum clock in $1656$ at the
Hague, applying Galileo's small-amplitude isochronism observation
to a practical timekeeper. The first Huygens clock kept time to
$\sim 10 \,\si{\second}$ per day, a fractional uncertainty of
$\sim 10^{-4}$. By $1675$, with the cycloidal cheek correction
(Huygens's own modification to maintain isochronism at finite
amplitude), accuracy reached $\sim 1 \,\si{\second}$ per day,
fractional uncertainty $\sim 10^{-5}$.

John Harrison's marine chronometer H4 (sea trial $1761$-$1762$)
solved the longitude problem by carrying mechanical time accurate
to roughly $5 \,\si{\second}$ over a $46$-day Atlantic crossing,
or $\sim 10^{-6}$ fractional accuracy under shipboard conditions:
temperature swings of $10$-$20 \,\si{\kelvin}$, sustained pitch
and roll, no possibility of attendant care. The H4 ran a
diamond-and-ruby escapement with a temperature-compensating
bimetal balance and was a hand-built artefact rather than a
manufactured object. Its commercial successors, mass-produced
through the nineteenth century, brought marine chronometry to
working navigators \cite{hist:ch06t3-sobel-longitude-1995}.

The Shortt-Synchronome free-pendulum clock ($1921$, William Shortt
at the Synchronome Co.) housed the master pendulum in an
evacuated tank, decoupled it from the impulse mechanism via a
slave pendulum, and reached fractional accuracies of $\sim 10^{-8}
$ ($\sim 10 \,\si{\milli\second}$ per day). The Shortt-Synchronome
was the time standard at the Royal Greenwich Observatory and at
NIST's predecessor laboratories into the $1940$s, when it was
displaced by quartz-crystal oscillators that surpassed it within
two decades of their introduction.

\subsection{The quartz era: Marrison and the wristwatch}

Warren Marrison at Bell Telephone Laboratories ($1927$) built the
first practical quartz-crystal clock, exploiting the
piezoelectric resonance of a thin quartz plate to maintain a
stable oscillation at $50 \,\si{\kilo\hertz}$
\cite{paper:ch06t3-marrison-1948}. The Marrison clock reached
fractional accuracies of $\sim 10^{-8}$, comparable to the
Shortt-Synchronome but at a fraction of the cost, the size of a
laboratory bench, and with no mechanical wear surfaces. Within a
decade, quartz oscillators were the working frequency reference
of radio broadcasting, telephone networks, and military timing
systems.

The decisive step toward mass adoption came in $1969$, when Seiko
released the Astron quartz wristwatch at the Tokyo Watch Fair.
The Astron used a $8{,}192 \,\si{\hertz}$ quartz resonator (later
$32{,}768 \,\si{\hertz} = 2^{15}$, the modern standard) driven by
a battery-powered integrated circuit. Within $20$ years, quartz
watches had displaced mechanical watches in the consumer market;
the wristwatch's fractional accuracy of $\sim 10^{-6}$ ($\sim 30
\,\si{\second}$ per year) was orders of magnitude better than a
$1968$ mechanical wristwatch and met the everyday-use accuracy
target. The same resonator and circuit, in $2026$, sit inside
every smartphone, every microcontroller, and every digital
appliance the reader encounters: the quartz crystal is the most
widely deployed precision instrument in human history.

\subsection{The atomic era: Essen and Parry to NIST-F1}

Louis Essen and Jack Parry at the National Physical Laboratory
built the first operational caesium-beam frequency standard in
$1955$, locking the microwave cavity frequency to the $9{,}192
{,}631{,}770 \,\si{\hertz}$ hyperfine transition of caesium-133
\cite{paper:ch06t3-essen-parry-1955}. The Essen-Parry standard
reached fractional accuracies of $\sim 10^{-10}$, two orders of
magnitude better than the best contemporary quartz oscillators.
The $1967$ CGPM redefined the second in terms of the caesium
transition, transferring the SI definition from astronomical
observation (the ephemeris second, derived from Earth's orbital
motion) to laboratory physics. The redefinition meant that a
laboratory could realise the SI second directly with a primary
standard, no astronomical observations required.

The caesium-beam architecture was succeeded by the caesium-fountain
architecture in the $1990$s. The fountain launches laser-cooled
caesium atoms upward, lets them traverse a microwave cavity twice
(once on the way up, once on the way down), and reads the
resulting atomic state population. The longer interrogation time
gives a narrower resonance and lower fractional uncertainty.
NIST-F1 (operational $1999$) reached $\sim 10^{-15}$ fractional
accuracy; NIST-F2 ($2014$) reached $\sim 10^{-16}$
\cite{web:nist-cesium-fountain}. The fountains are the present
primary frequency standards at most major national metrology
laboratories.

\subsection{The optical era: lattice clocks and the future SI second}

Optical frequency standards, lock to electronic transitions at
hundreds of terahertz rather than the microwave hyperfine
transitions of caesium. The higher transition frequency gives a
higher line quality factor and a narrower resonance, extending
fractional accuracies into the $10^{-18}$ range. The strontium
optical lattice clock at NIST/JILA reached fractional accuracy
$\sim 4 \times 10^{-18}$ in $2014$ and improved further in
subsequent years \cite{web:ch06t3-nist-strontium-2024}. The
optical lattice traps thousands of atoms in a standing-wave laser
field at a wavelength chosen so the trap shifts the clock
transition by exactly zero (the magic wavelength); the trapped
atoms are interrogated by a probe laser whose frequency is
stabilised to the atomic transition via Pound-Drever-Hall
locking.

Optical clocks remain laboratory systems in $2026$: a working
strontium lattice clock fills a room with lasers, vacuum chambers,
frequency combs, and environmental controls. The optical
frequency comb (Th\'eodor H\"ansch and John Hall, Nobel $2005$)
provides the bridge between the optical transition (at $\sim 429
\,\si{\tera\hertz}$ for strontium) and an electronic counter at
microwave or radio frequencies. The combs run as femtosecond
mode-locked lasers whose output is a coherent spectrum of equally
spaced frequency lines spanning an octave; the line spacing
equals the laser's pulse repetition rate, measurable with an
electronic counter, and the absolute frequency of any line is
known if one line is anchored to a reference (typically the
caesium standard, or another optical clock).

The CCTF and national metrology laboratories are preparing the
technical basis for a future optical redefinition of the SI
second. The schedule should be cited to current BIPM or CCTF
material rather than fixed in this text; the working point is
that the SI second is on a trajectory toward an optical
redefinition once the international comparison campaign matures.

\subsection{NIST-on-a-chip: precision at the milliwatt scale}

Parallel with the laboratory optical clock effort, a NIST
programme called \engterm{NIST-on-a-Chip} has miniaturised the
caesium frequency standard onto a circuit board
\cite{web:ch06t3-nist-on-a-chip-2024}. The Microsemi (now
Microchip Technology) SA.45s chip-scale atomic clock, in
production since $2011$, uses coherent population trapping in a
vapour cell of caesium-133, draws $\sim 120 \,\si{\milli\watt}$,
occupies $\sim 35 \,\si{\centi\meter\cubed}$, and reaches
$\sigma_{y}(\tau = 1 \,\si{\day}) \approx 10^{-11}$
\cite{web:ch06t3-microsemi-csac-2024}. Fractional accuracy is
$\sim 10^{-10}$, three orders of magnitude better than the best
TCXO and three orders of magnitude worse than a caesium fountain.
The portability is the point: a CSAC fits in a hand, runs on a
small battery for hours, and provides atomic-class holdover for
mobile applications (military timing, undersea sensors,
high-frequency trading, distributed phasor measurement units).

The NIST-on-a-Chip programme extends to other quantities (the
silicon photonic optical-frequency reference, the on-chip Kibble
balance, the chip-scale thermometer). The pattern is the same as
the historical arc: take a laboratory-grade primary standard,
miniaturise it through engineering, distribute it across working
measurement chains. The miniaturisation trades absolute accuracy
for portability and cost; the working engineer chooses the
appropriate operating point on the Pareto curve.

\subsection{Why time anchored everything else}

The arc has a structural feature: each new clock improved
fractional uncertainty by roughly an order of magnitude, and at
each step, working measurement chains began to express other
quantities through time. The metre was redefined in $1983$ as
the distance light travels in $1/299{,}792{,}458 \,\si{\second}$,
transferring length precision from the krypton-86 wavelength
standard to the caesium frequency standard. The volt, through the
Josephson effect, is expressed as a frequency. The ampere, through
the $2019$ redefinition, is expressed through the elementary
charge and frequency-defined units. The kilogram, through the
Kibble balance, is expressed through the Planck constant and
electrical quantities that ground out in frequency. By $2026$, the
SI is structurally a frequency-anchored system: the second is the
primary reference, and the other six base units inherit their
precision from it.

The lesson for working engineers, even those not in metrology,
is that the time signal in any measurement chain is the cleanest
reference available. When a measurement can be expressed as a
time interval times a calibrated rate, the rate's fractional
uncertainty sets the floor, and modern frequency references put
that floor below most other measurands' achievable precision. The
discipline of pulling a measurement back to a time interval, when
the physics allows, is a working pattern the reader will see
repeated through the rest of \emph{Engineering}.

\section{Failure: clock drift and the Patriot missile}\pathtag{core}

Clock errors fail measurement systems through accumulation: a
small, slowly accumulating timing error can grow into an
operationally significant error if the system runs continuously
without resetting. The canonical case is the Patriot missile
battery at Dhahran, Saudi Arabia, on 25 February 1991.

\subsection{The Patriot at Dhahran, 1991}

A US Army Patriot missile battery at Dhahran, Saudi Arabia, failed
to intercept an incoming Iraqi Scud missile on 25 February 1991,
killing 28 US service members and wounding approximately 100
others. The U.S.\ General Accounting Office investigated the
incident under congressional request and traced the failure to a
software-clock arithmetic error that had accumulated to
approximately $0.34 \,\si{\second}$ over $100 \,\si{\hour}$ of
continuous operation, large enough that the Patriot's range gate
computed the Scud's predicted position outside the search window
and the system did not engage \cite{acc:gao-patriot-1992}. The
canonical record of the case is the registry entry at
\repopath{docs/research/accidents/patriot-dhahran-1991.md}; the
account here paraphrases that entry and the GAO report, paragraph
by paragraph.

\paragraph{Technical mechanism.}
The Patriot's tracking software represented elapsed time as the
number of tenths of a second since the system was started, stored
as an integer counter. When the system needed to convert this
counter to seconds for the range-gate calculation, it multiplied
by a binary representation of $1/10$. The exact decimal $0.1$ has
no exact binary representation; truncating it to a $24$-bit
fixed-point fraction introduces a per-conversion error of
approximately $9.5 \times 10^{-8}$ in the converted seconds value.
The error accumulated linearly with the integer counter
\cite{acc:gao-patriot-1992}. After approximately $100 \,\si
{\hour}$ of continuous operation, the integer counter had reached
approximately $3.6 \times 10^{6}$ tenths of a second, and the
cumulative conversion error had reached approximately $0.34 \,\si
{\second}$. The Patriot's range-gate calculation used the
converted time to predict where an incoming target would be at the
next radar look. A target moving at the Scud's speed of
approximately $1{,}676 \,\si{\meter\per\second}$ over $0.34 \,\si
{\second}$ travels approximately $570 \,\si{\meter}$. The
Patriot's predicted position was therefore offset from the actual
position by roughly $570 \,\si{\meter}$, larger than the range
gate, and the system reported ``no track'' rather than engaging
\cite{acc:gao-patriot-1992}. The mechanism is the same pattern a
reader will see in long-running measurement chains:
representation error, linear accumulation, threshold crossing, and
late correction. \Cref{fig:vol01:ch06:patriot-drift} plots the
cumulative-error ramp and marks the range-gate threshold.

\input{volumes/01-quantity/ch06-time-frequency-signals/figures/fig-patriot-drift.tex}

\paragraph{Organisational context.}
The GAO report identifies contributing institutional factors
beyond the technical error \cite{acc:gao-patriot-1992}. Patriot
batteries were not designed to operate continuously for one
hundred hours; the operational concept assumed periodic restarts
that would reset the time counter. Field operators in the Gulf War
deployment had been instructed to keep the systems running
continuously to maintain readiness. The accumulated-error effect
had been observed in field tests and reported through the chain,
but the patch and the operational-restart guidance moved through
different distribution channels and arrived at field units
inconsistently. A software patch correcting the conversion error
had been developed and was being distributed to Patriot batteries
at the time of the Dhahran incident; the Dhahran battery's patch
arrived on 26 February 1991, the day after the incident
\cite{acc:gao-patriot-1992}. The institutional pattern is a
mismatch between two distribution channels: a technical patch on
one timetable, an operational-restart guidance on another, and a
deployed unit that received the guidance to operate continuously
before it received the patch that made continuous operation safe.

\paragraph{Lesson for timing systems.}
A timing system whose internal counter accumulates a per-step
error must be analysed against its expected continuous-operation
duration. A counter that increments at $10 \,\si{\hertz}$ and
introduces $9.5 \times 10^{-8} \,\si{\second}$ per increment
accumulates one second of error in about $30 \,\si{\hour}$ of
continuous operation; the precise number for a given system is the
relevant design quantity. A timing-system design that quietly
assumes restarts at intervals shorter than the drift threshold
must publish that assumption in the operating manual, in the
qualification testing, and in the patch-distribution channel that
adds the assumption to deployed units. A timing system that
crosses an unannounced drift threshold during operation is a
clock-failure pattern that a Chapter 6 reader is expected to
recognise on inspection. The reader's working response, after this
chapter, is to ask three questions of any long-running timing
system: what is the per-step arithmetic error, what is the
expected continuous-operation interval, and what reset mechanism
prevents the per-step error from accumulating past the application
threshold? The Patriot case is read in this chapter as the
canonical timing-systems failure; Volume VII Chapter 2 returns to
the same case with the number-representation discipline that would
have prevented the per-step error, and Volume X takes up the
organisational lesson at the institutional scale.

\subsection{The mechanism abstracted}

The Patriot mechanism has the same structure as a long catalogue of
clock-drift failures across domains. A small per-step error
accumulates linearly with the number of steps; if the step counter
is unbounded, so is the accumulated error. A long-running
measurement, control, or analysis system that does not reset its
clock periodically therefore inherits an unbounded drift unless
the per-step error is exactly zero (which, in floating-point
arithmetic, it almost never is).

The same pattern appears in:
\begin{itemize}[leftmargin=*]
  \item Inertial navigation systems whose accelerometer biases
    integrate into velocity errors and whose gyroscope biases
    integrate into orientation errors. The errors grow with the
    duration of unaided operation; aircraft and ships periodically
    update their inertial systems against external references
    (GPS, celestial, terrain).
  \item Numerical integration of differential equations in
    long-running simulations, where roundoff accumulates step by
    step. The discipline of bounded-roundoff time integration is
    a routine concern in scientific computing.
  \item Financial accounting systems whose internal counters drift
    against an external reference (an exchange's clock, a national
    central bank's settlement clock); the resulting timestamps can
    misorder transactions across systems.
  \item Power-grid phasor measurement units whose local clocks
    must remain synchronised to UTC at the microsecond level for
    the resulting phasor measurements to be comparable across the
    grid; clock drift between PMUs corrupts the wide-area
    measurement record.
\end{itemize}

A second canonical case, briefer than the Patriot but of the same
family, is the GPS week-number rollover. The legacy GPS C/A
navigation message carries the week number in a 10-bit field, so
the broadcast value wraps from 1023 to 0 every $2^{10} = 1024$
weeks, approximately 19.7 years. Rollovers occurred at the end of
21 August 1999 and the end of 6 April 2019. A receiver reconstructs
the absolute week by combining the broadcast value with a pivot
date held in firmware (typically the compile date or the last
firmware-update date); a receiver whose pivot has not been
refreshed before a rollover may compute a calendar date that is an
integer multiple of 1024 weeks too early. The 2019 rollover
produced documented timing-system outages and erroneous time
stamps in installations whose firmware-update path had lapsed; the
position service of a navigation-only receiver is not directly
affected, because the within-week time-of-week field carries the
ranging information, but time-transfer and time-stamp applications
are exposed even when position outputs appear nominal
\cite{web:gpsgov-week-rollover-2019,web:dhs-cisa-gps-rollover-2019,
web:nist-gps-week-rollover-2019}. The newer CNAV message uses a
13-bit week field and does not roll over until 2137. The registry
entry at \repopath{docs/research/accidents/gps-week-rollover-1999-2019.md}
collects the technical and organisational mechanisms; the two
canonical responses, parallel CNAV-capable receivers in critical
timing chains and explicit holdover plans for the rollover dates,
match the Patriot's discipline list one for one.

\begin{principle}[Long-running measurement systems require periodic
clock reset against a stable reference]
A measurement system that uses an internal counter to represent
time, and that runs continuously, accumulates a drift whose
magnitude depends on the per-step arithmetic, the per-step
calibration, and the run length. The drift becomes operationally
significant when it crosses a threshold set by the application, not
by the engineer's intuition. The discipline that bounds the drift
is to reset or recalibrate the internal time representation against
a stable external reference at intervals shorter than the
application's drift threshold. The Patriot case illustrates the
failure mode in which the design assumed periodic restarts, the
operational concept did not enforce them, and the patch arrived
after the incident.
\end{principle}

\subsection{The discipline}

The engineering response to clock-drift failure is layered:

\begin{enumerate}[leftmargin=*]
  \item Choose a time representation whose per-step arithmetic
    error is either exactly zero (integer counters with exactly
    representable scale factors) or bounded and characterised
    (integer counters with documented per-step rounding).
  \item Document the operational concept of the system: how long
    it runs continuously, what restarts are scheduled, what the
    drift accumulation is over the longest expected continuous
    run.
  \item Build periodic reset against a stable external reference
    into the system's operational procedure. The reference can be
    UTC via GPS, a network time service, a local atomic clock, or
    the application's own zero-point markers (start-of-day,
    start-of-cycle).
  \item Test the system at the longest expected continuous run.
    Drift is a long-time-scale failure mode; a short qualification
    test can pass while the fielded system still fails after days
    or weeks of continuous operation.
  \item When a patch corrects a drift mechanism, distribute the
    patch and the operational guidance through the same channel,
    so a deployed unit cannot have one without the other.
\end{enumerate}

The Patriot case had a software patch and an operational guidance
on continuous-operation duration; the two distribution channels
were not synchronised, and the Dhahran battery had received the
guidance to operate continuously without yet receiving the patch
that made continuous operation safe. The discipline of joint
distribution is the institutional lesson, taken up in detail in
Volume X.

\section{Project}\pathtag{core}

\begin{project}[Twenty-four-hour mains-frequency record]
The reader records or estimates the mains frequency at their
location for a continuous twenty-four-hour period, at one-minute
resolution or finer, with an estimated standard uncertainty per
sample, and compares the resulting record to the grid operator's
published frequency record where one is available. Per Q55-ch06 the
project runs as a household plus analysis hybrid track at low
hazard.

\textbf{Track}. Household plus analysis (hybrid). The preferred
route is optical: a photodiode or a smartphone camera observes the
flicker from a mains-powered lamp. The electrical route is allowed
only with a sealed, certified plug-in AC-output adapter or another
certified isolated low-voltage source; the reader does not open,
modify, or wire mains equipment. If a safe certified isolated
adapter is not already available, the optical route is mandatory.

\textbf{Hazard class}: low under the constraints above. The reader
uses an optical coupling or a certified plug-in adapter, never a
direct galvanic connection to mains-voltage wiring. The smartphone
or laptop performing the recording remains on its own battery or
on its own properly isolated charger; the optical or adapter
coupling to the mains is one-way on the receiver side. The reader
does not improvise a transformer, splice into mains conductors, or
open mains-powered equipment.

\textbf{Tools}. A mains-powered LED or fluorescent lamp whose
flicker is visible in a short test recording (LED bulbs typically
flicker at twice the mains frequency, $100$ or $120 \,\si{\hertz}$,
with depth depending on the driver); a smartphone camera or
photodiode for the optical route, or a certified isolated
low-voltage adapter for the electrical route; a recorder or logger
(smartphone, laptop, or microcontroller); and software for
zero-crossing detection, sinusoid fitting, or peak-frequency
estimation (a free audio tool such as Audacity, a short Python
script using SciPy or NumPy, or a spreadsheet for post-processing).
The durable requirement is a timestamped frequency estimate every
minute. The reader may store derived zero-crossing times or one
short window per minute instead of storing a continuous raw
24-hour audio or video file. Optionally, an oscilloscope for
verifying the acquisition chain before the long recording starts.

\textbf{Procedure}.

\begin{enumerate}[leftmargin=*]
  \item \emph{Plan the measurement chain}. Decide whether the
    record will be continuous raw data, short windows collected once
    per minute, or derived zero-crossing timestamps. Estimate
    storage, battery life, heat, and sample-clock accuracy before
    starting; a continuous 16-bit $44{,}100 \,\si{\hertz}$ WAV file
    for 24 hours is roughly $7.6$ GB, and most smartphones cannot
    run a continuous slow-motion video for that duration. The
    sample rate must be high enough for the chosen estimator, and
    the anti-aliasing or front-end bandwidth must be documented.

  \item \emph{Verify the chain on a short recording}. Before the
    long recording, take a five-minute test recording. Inspect the
    waveform; verify that a clean periodic signal is present at
    $50 \,\si{\hertz}$ (or $60 \,\si{\hertz}$ if applicable) for an
    audio-coupled recording, or at twice the mains frequency for an
    optical recording from an LED lamp. Compute the zero-crossing
    rate or the peak FFT frequency and confirm it matches the
    expected nominal frequency to within $\sim 0.1 \,\si{\hertz}$.

  \item \emph{Run the 24-hour recording}. Start the recording at a
    documented local time. Note the recording-device clock's
    relationship to UTC (most smartphones and laptops are
    synchronised through network time service to within tens of
    milliseconds). Do not stop or restart the recording during the
    24-hour window unless an interruption is unavoidable; if it
    becomes unavoidable, log the start and end of each segment.

  \item \emph{Analyse in one-minute windows}. Divide the recording
    into one-minute segments. For each segment, estimate frequency
    by zero-crossing detection with interpolation, by sinusoid
    fitting, or by an FFT method with peak interpolation or phase
    tracking. Do not report the largest raw FFT bin as a millihertz
    estimate; its spacing is $1/60 \,\si{\hertz} \approx 16.7 \,\si
    {\milli\hertz}$ for a one-minute window. Record the estimated
    frequency and an estimated standard uncertainty.

  \item \emph{Construct an uncertainty budget}. List each
    uncertainty source: sample-clock fractional frequency error,
    estimator error (zero-crossing interpolation, sinusoid fit, or
    FFT peak), waveform distortion, noise, front-end delay
    variation, and any drift in the optical or adapter coupling
    over the 24-hour window. If the sample clock is not disciplined
    to a known reference, estimate or measure its fractional error
    before claiming millihertz-level frequency accuracy. Propagate
    the components to a combined standard uncertainty per the
    rules of Chapter 4.

  \item \emph{Compare to a published reference}. Many grid
    operators publish the grid frequency at one-minute or finer
    resolution. In Europe, the ENTSO-E network publishes
    grid-frequency data; in the United Kingdom, the National Grid
    publishes its frequency record; in North America, several
    regional ISOs publish their frequency. Identify the published
    record applicable to the reader's grid, download it for the
    same 24-hour window, and compare. Plot both records on the
    same axes and compute the residual.

  \item \emph{Plot and tabulate}. Plot the reader's frequency
    record over 24 hours, with error bars at each sample. Plot
    the published frequency record on the same axes. Plot the
    residual (reader minus published) versus time. Tabulate the
    mean, standard deviation, and time-of-day extremes for both
    records.
\end{enumerate}

\textbf{Record}. The deliverable includes: the recording chain
description (instrument, coupling, software, file format); the
five-minute verification result; the 24-hour frequency record at
one-minute resolution with per-sample uncertainty; the uncertainty
budget; the comparison to the published reference; the residual
plot.

\textbf{Reflection (1500 words)}. The reader writes a 1500-word
narrative addressing:

\begin{itemize}[leftmargin=*]
  \item Which uncertainty source dominated the per-sample standard
    uncertainty, and why.
  \item Whether a time-of-day pattern is visible in the frequency
    record (load shifts in morning, evening, overnight) and what
    physics would account for it.
  \item Whether the reader's record and the published record agree
    within their combined uncertainties at every minute, or whether
    a systematic offset or drift is visible.
  \item What would change in the reader's confidence if the
    application of the measurement were a synchronous mains-driven
    clock (fractional time error after 24 hours), a power-grid
    phasor measurement (microsecond-level synchronisation
    requirement), or a frequency-stability claim for a household
    appliance (millihertz-level resolution).
  \item One environmental factor (the recording-device clock's
    drift, the audio adapter's group delay, the lamp's driver
    behaviour, the laboratory mains harmonic content) that the
    project did not characterise but a real engineering
    deployment would.
\end{itemize}

\textbf{Deliverable}. The recording chain description, the
verification, the 24-hour data, the uncertainty budget, the
comparison plot, the residual analysis, and the reflection. The
general editor's reference 24-hour record from a representative
European and North American location is published online; the
reader is encouraged to compare \emph{after} completing their own.

\textbf{Time}. Setup and verification: one to two hours. Recording:
24 hours of unattended operation. Analysis and plotting: three to
six hours. Reflection: two to three hours. Total reader time
approximately six to eleven hours, plus the unattended recording
window.

\textbf{Phases and deliverables (recap)}. The project decomposes
into five phases, each with an explicit deliverable that can be
checked before the next phase begins:

\begin{enumerate}[leftmargin=*]
  \item \emph{Plan and verify} (Procedure steps 1-2). Deliverable:
    a one-page measurement-chain description plus the five-minute
    verification record showing the expected periodic signal
    within $\sim 0.1 \,\si{\hertz}$ of nominal.
  \item \emph{Record} (Procedure step 3). Deliverable: a 24-hour
    recording (continuous, one-window-per-minute, or derived
    timestamps) with documented start time and recorder clock
    relationship to UTC.
  \item \emph{Estimate per window} (Procedure step 4). Deliverable:
    a 1440-row table of (UTC timestamp, frequency estimate,
    per-sample uncertainty).
  \item \emph{Budget} (Procedure step 5). Deliverable: an uncertainty
    budget table listing each source, its propagated contribution,
    and the combined standard uncertainty.
  \item \emph{Compare and report} (Procedure steps 6-7 plus
    Reflection). Deliverable: the two-record plot, the residual
    plot, the tabulated mean and standard deviation for both
    records, and the 1500-word reflection.
\end{enumerate}

\textbf{Success criteria}. The project counts as successful if
(a) the recording chain is documented in enough detail that a peer
could reproduce the measurement, (b) the per-sample uncertainty is
defended against the dominant source rather than asserted, (c) the
residual against the published reference is at the millihertz
level over most of the 24 hours (with explanations for any
excursions), and (d) the reflection takes an explicit position on
which uncertainty source dominated and how a future deployment
would change.

\textbf{Worked example}. A reference implementation of the
zero-crossing-with-interpolation estimator (Procedure step 4) is on
disk at \texttt{code/mains\_zero\_crossing.py}; it generates a
synthetic 60-second $50.01 \,\si{\hertz}$ signal at $10 \,\si
{\kilo\hertz}$, runs the estimator over one-second windows, and
prints the estimator's mean and standard deviation. A reader who
adapts the script to read their own recording's WAV or CSV file
inherits the algorithm and the seed for the noise-floor sanity
check. The companion FFT-based estimator with parabolic peak
interpolation is at \texttt{code/fft\_noisy\_signal.py}; comparing
the two estimators on the same recording is the simplest way to
catch estimator-specific bias. The data file
\repopath{data/sample_rates.csv} carries the working table of
sample rates and bandwidths that informs the chain-design step.

\textbf{Failure modes to watch for}. A recording-device clock
running fast or slow by $10^{-5}$ produces a systematic frequency
offset of about $5 \times 10^{-4} \,\si{\hertz}$ on a $50 \,\si
{\hertz}$ signal, half a millihertz, comparable to the target
uncertainty. A lamp driver whose flicker is not exactly at twice
the mains frequency (some modern LED drivers regulate at higher
frequencies and modulate at a slower envelope) produces a record
whose frequency content is harder to interpret than expected. An
optical recording made through a non-anti-aliased smartphone video
pipeline can produce aliasing artefacts in the recorded
brightness. Each is a candidate for the reflection's dominant
uncertainty source.
\end{project}

\section{Exercises}

The exercises begin with period-frequency arithmetic and end with
sampling diagnosis, GPS error budgets, and clock-drift failure
analysis.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A signal has frequency $f = 60 \,\si{\hertz}$. Compute its period in
milliseconds and its angular frequency in radians per second.

\paragraph{Worked solution.}
\textbf{Set-up.} A sinusoidal signal of cyclic frequency $f$ has
period $T = 1/f$ (the inverse of cycles per second is seconds per
cycle) and angular frequency $\omega = 2\pi f$ (each cycle is
$2\pi$ radians). The three quantities encode the same information
on different scales; carrying all three in working memory avoids
the recurring $2\pi$ slip that appears in phase calculations.

\textbf{Given.} $f = 60 \,\si{\hertz}$, the North American mains
frequency.

\textbf{Step 1: compute the period.} $T = 1/f = 1/60 \,\si{\second}
= 0.01\overline{6} \,\si{\second} = 16.67 \,\si{\milli\second}$.
The decimal representation $16.\overline{6} \,\si{\milli\second}$
is exact; the truncation to $16.67$ introduces about
$2 \times 10^{-4}$ relative error.

\textbf{Step 2: compute the angular frequency.} $\omega = 2\pi f =
2\pi \times 60 = 120\pi \,\si{\radian\per\second}$. Numerically,
$120\pi \approx 376.99 \,\si{\radian\per\second}$, often
remembered as $\omega \approx 377 \,\si{\radian\per\second}$.

\textbf{Step 3: cross-check via the radian-period product.}
$\omega T = 2\pi$ (one period covers one full cycle in radians);
numerically $377 \times 0.01667 = 6.28 \approx 2\pi$, confirming
the consistency.

\textbf{Sanity check.} The European mains at $50 \,\si{\hertz}$
has period $20 \,\si{\milli\second}$ and $\omega = 100\pi \approx
314 \,\si{\radian\per\second}$. A reader who carries both
$120\pi$ and $100\pi$ for the two grids has a working dictionary
that translates between the two systems without arithmetic.

\textbf{Common error.} Dropping the factor of $2\pi$ when moving
between $f$ in hertz and $\omega$ in radians per second. The
working remedy is to write the conversion explicitly the first
time it appears in any derivation: $\omega = 2\pi f$, not
$\omega = f$. The same slip appears at the interface between
discrete-time DFT bin indices and continuous frequencies and is
worth flagging again there.
\end{exercise}

\begin{exercise}
A clock with fractional accuracy $10^{-6}$ keeps time over one year.
Compute the maximum cumulative error in seconds. Compare to the
$0.34 \,\si{\second}$ Patriot accumulated error after $100 \,\si
{\hour}$ of operation; identify the fractional accuracy implied by
the Patriot's per-conversion error.

\paragraph{Worked solution.}
\textbf{Set-up.} Fractional clock accuracy $\Delta f/f$ converts
to cumulative timing error $\Delta t = (\Delta f/f) \times t$ over
an elapsed interval $t$. A $10^{-6}$ fractional error accumulates
one microsecond per second, one millisecond per thousand seconds,
and so on. The conversion is the working bridge between
specification (a fractional figure that appears on the datasheet)
and operational performance (a number of seconds that matter
against a system's timing budget).

\textbf{Given.} Fractional accuracy $\Delta f/f = 10^{-6}$; year
elapsed time $\approx 3.156 \times 10^{7} \,\si{\second}$.
Comparison data: Patriot system accumulated $\Delta t = 0.34 \,\si
{\second}$ over $t = 100 \,\si{\hour}$ of operation
\cite{acc:gao-patriot-1992}.

\textbf{Step 1: cumulative error over one year.}
$\Delta t_{\text{yr}} = 10^{-6} \times 3.156 \times 10^{7} =
31.6 \,\si{\second}$. The figure matches the rule-of-thumb
$\approx 32 \,\si{\second}/\text{yr}$ band for wristwatch-grade
quartz oscillators.

\textbf{Step 2: convert the Patriot operational data to a
fractional figure.} The Patriot operated for $t = 100 \,\si{\hour}
= 3.6 \times 10^{5} \,\si{\second}$ and accumulated $\Delta t =
0.34 \,\si{\second}$, giving $\Delta f/f = 0.34/(3.6 \times 10^{5})
\approx 9.4 \times 10^{-7}$, just under $10^{-6}$.

\textbf{Step 3: compare against the wristwatch.} The Patriot
fractional figure is marginally better than a typical wristwatch.
The two are in the same order of magnitude despite the Patriot
being an air-defence system and the wristwatch a consumer good;
the fractional figure alone does not capture the operational
significance.

\textbf{Sanity check.} The Patriot's $0.34 \,\si{\second}$ at
$100 \,\si{\hour}$ scales linearly: a $1000 \,\si{\hour}$ mission
would accumulate $3.4 \,\si{\second}$, and a $10 \,\si{\hour}$
mission would accumulate $34 \,\si{\milli\second}$. The
operational threshold for missing the target (a few hundred
milliseconds) is reached after about $50 \,\si{\hour}$ of
continuous operation, consistent with the GAO report's account
of which Patriot batteries were and were not vulnerable
\cite{acc:gao-patriot-1992}.

\textbf{Common error.} Treating fractional accuracy as a quality
score. The Patriot case shows that an acceptable fractional
figure can produce an operationally catastrophic error if the
system runs continuously past the threshold at which the
cumulative error matters. The correct working pattern is to
convert the fractional figure into a cumulative error against the
system's mission length, then compare against the mission's
timing tolerance. The system specification should bound the
maximum continuous operating time as a function of fractional
clock accuracy and timing tolerance, not the fractional accuracy
alone.
\end{exercise}

\begin{exercise}
A digital recording samples at $f_{s} = 48 \,\si{\kilo\hertz}$ for
$N = 65{,}536$ samples. Compute the recording duration and the
frequency resolution of a single FFT computed over the entire
recording.

\paragraph{Worked solution.}
\textbf{Set-up.} A discrete Fourier transform over $N$ samples
acquired at rate $f_{s}$ produces $N$ frequency bins linearly
spaced from $0$ to $f_{s}$. The recording duration is
$\tau = N/f_{s}$ and the bin spacing is $\Delta f = f_{s}/N = 1/
\tau$. The reciprocal relation between bin spacing and recording
duration is the working dictionary for designing FFT-based
measurements.

\textbf{Given.} $f_{s} = 48 \,\si{\kilo\hertz}$ (a standard
audio-grade sample rate), $N = 65{,}536 = 2^{16}$ samples (the
power-of-two length convenient for the FFT).

\textbf{Step 1: recording duration.} $\tau = N/f_{s} = 65{,}536/
48{,}000 \,\si{\second} \approx 1.3653 \,\si{\second}$.

\textbf{Step 2: frequency bin spacing.} $\Delta f = f_{s}/N =
48{,}000/65{,}536 \,\si{\hertz} \approx 0.7324 \,\si{\hertz}$, or
equivalently $1/\tau \approx 0.7324 \,\si{\hertz}$, confirming
the reciprocal relation.

\textbf{Step 3: extract the working dictionary.} To resolve two
tones separated by $\Delta f_{\text{target}}$, the recording
duration must satisfy $\tau \geq 1/\Delta f_{\text{target}}$,
independent of the sample rate. For $\Delta f_{\text{target}} =
0.1 \,\si{\hertz}$, $\tau \geq 10 \,\si{\second}$; the sample
rate sets the highest frequency that can appear (Nyquist), not
the resolution at lower frequencies.

\textbf{Sanity check.} A two-tone test at $440 \,\si{\hertz}$ and
$440.7 \,\si{\hertz}$ on this $65{,}536$-sample recording would
straddle one bin (at $\Delta f = 0.73 \,\si{\hertz}$ bin spacing,
the two tones fall within adjacent bins). A test at $440$ and
$441 \,\si{\hertz}$, with $1 \,\si{\hertz}$ separation, exceeds
$\Delta f$ and the two peaks are visibly resolved.

\textbf{Common error.} Believing that a higher sample rate
improves frequency resolution. The sample rate sets the Nyquist
ceiling; the recording duration sets the resolution. A reader
who tries to resolve a $0.1 \,\si{\hertz}$ separation with a
$1 \,\si{\second}$ recording at $1 \,\si{\mega\hertz}$ sample
rate still sees a $1 \,\si{\hertz}$ bin spacing. The fix is more
samples in time, not more samples per second.
\end{exercise}

\begin{exercise}
A signal at $f = 7{,}000 \,\si{\hertz}$ is sampled at $f_{s} = 8{,}000
\,\si{\hertz}$. State whether the Nyquist condition is satisfied. If
not, compute the alias frequency that appears in the sampled record.

\paragraph{Worked solution.}
\textbf{Set-up.} The Nyquist condition asserts that a band-limited
signal of maximum frequency $f_{\text{max}}$ can be uniquely
recovered from samples taken at rate $f_{s}$ if and only if
$f_{s} > 2 f_{\text{max}}$. When a signal at $f$ exceeds the
Nyquist frequency $f_{s}/2$, its samples are indistinguishable
from those of an alias at $f_{\text{alias}} = |f - n f_{s}|$, with
$n$ the integer that places the alias in $[0, f_{s}/2]$.

\textbf{Given.} Signal frequency $f = 7{,}000 \,\si{\hertz}$;
sample rate $f_{s} = 8{,}000 \,\si{\hertz}$. Goal: state whether
Nyquist holds; if not, compute the alias.

\textbf{Step 1: check the Nyquist condition.}
$f_{s}/2 = 4{,}000 \,\si{\hertz}$; $f = 7{,}000 \,\si{\hertz}$;
$f > f_{s}/2$, so Nyquist is violated. Aliasing will occur.

\textbf{Step 2: choose $n$ and compute the alias.} The integer
$n$ that places the alias in $[0, f_{s}/2]$ is the integer
closest to $f/f_{s}$. Here $f/f_{s} = 7{,}000/8{,}000 = 0.875$,
so $n = 1$. The alias is $f_{\text{alias}} = |7{,}000 - 1 \times
8{,}000| = 1{,}000 \,\si{\hertz}$.

\textbf{Step 3: confirm the alias sits in $[0, f_{s}/2]$.}
$1{,}000 < 4{,}000$, so the alias lies in the principal Nyquist
zone. The sampled record will report a clean $1{,}000 \,\si
{\hertz}$ sinusoid, unrecognisable as a degraded measurement of
the $7{,}000 \,\si{\hertz}$ original.

\textbf{Sanity check.} The script at
\texttt{code/aliasing\_demo.py} runs the same calculation for
both $7{,}000$ and $7{,}500 \,\si{\hertz}$ inputs at $f_{s} =
8{,}000 \,\si{\hertz}$ and verifies the alias numerically. A
second sanity check: the sum $f + f_{\text{alias}} = 7{,}000 +
1{,}000 = 8{,}000 = f_{s}$, which is the standard alias-pair
relation for a frequency just above Nyquist.

\textbf{Common error.} Assuming that satisfying Nyquist at the
\emph{nominal} signal frequency is sufficient. The Nyquist
condition must hold for all spectral content actually present in
the input, not just the nominal carrier. The anti-aliasing filter
exists to enforce the condition on the analog side; without it,
high-frequency noise and out-of-band interferers alias into the
working band and contaminate the measurement. The working
discipline is to specify the anti-aliasing filter as part of the
sampling chain, not as an optional extra.
\end{exercise}

\begin{exercise}
A GPS receiver measures the arrival time of a satellite signal with
standard uncertainty $5 \,\si{\nano\second}$. Compute the
corresponding standard uncertainty in the pseudorange. State the
assumption used (vacuum speed of light) and whether atmospheric
delay would worsen or improve the bound.

\paragraph{Worked solution.}
\textbf{Set-up.} A GPS receiver measures the time of arrival of
each satellite signal. The corresponding pseudorange is
$\rho = c \tau$, where $c$ is the propagation speed and $\tau$ is
the measured travel time. The pseudorange uncertainty is
$\sigma_{\rho} = c \,\sigma_{t}$ at the vacuum approximation,
plus separate atmospheric and multipath terms that the receiver
either models or absorbs into its position-solution residuals.

\textbf{Given.} Timing uncertainty $\sigma_{t} = 5 \,\si{\nano
\second}$ on a single pseudorange. Vacuum speed of light
$c = 2.998 \times 10^{8} \,\si{\meter\per\second}$, normally
rounded to $3 \times 10^{8}$ for back-of-envelope work. The
working conversion is $1 \,\si{\nano\second} \Leftrightarrow 30
\,\si{\centi\meter}$.

\textbf{Step 1: vacuum-approximation pseudorange uncertainty.}
$\sigma_{\rho} = c \sigma_{t} = 3 \times 10^{8} \times 5 \times
10^{-9} = 1.5 \,\si{\meter}$.

\textbf{Step 2: atmospheric correction.} The L1 single-frequency
ionospheric delay at zenith is typically $5$ to $15 \,\si{\meter}$
during the day and $1$ to $3 \,\si{\meter}$ at night, increasing
with elevation through the obliquity factor. The tropospheric
delay is approximately $2.3 \,\si{\meter}$ at zenith and scales
with elevation. The receiver's Klobuchar-model or dual-frequency
correction reduces residual ionospheric error to about $0.5$ to
$2 \,\si{\meter}$; the troposphere is corrected by a model based
on temperature, pressure, and humidity that closes residual to
$\sim 0.3 \,\si{\meter}$.

\textbf{Step 3: combined pseudorange budget.} The pseudorange
uncertainty after corrections is
$\sigma_{\rho,\text{total}} = \sqrt{(1.5)^{2} + (2.0)^{2} +
(0.3)^{2}} \approx 2.5 \,\si{\meter}$ for a single-frequency L1
receiver, dominated by residual ionospheric error rather than by
the receiver's timing measurement.

\textbf{Sanity check.} The pseudorange uncertainty of $2.5 \,\si
{\meter}$ converts through the geometric dilution of precision
(typically $1.5$ for an open-sky four-satellite fix) to a
horizontal position uncertainty of about $4 \,\si{\meter}$,
consistent with the open-sky horizontal accuracy reported by
consumer smartphones.

\textbf{Common error.} Reporting the vacuum-approximation
pseudorange uncertainty as if it were the working accuracy of the
position fix. The timing measurement is excellent; the
atmosphere is what limits the single-frequency receiver. A dual-
frequency receiver (L1 + L2 or L1 + L5) cancels the
first-order ionospheric term and approaches the vacuum bound;
modern smartphones (current as of 2026) increasingly carry
dual-frequency support for exactly this gain.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
For a sinusoidal signal $x(t) = A \cos(2\pi f t)$ delayed by a time
$\tau$ to produce $y(t) = x(t - \tau)$, show by direct substitution
that $y(t) = A \cos(2\pi f t - 2\pi f \tau)$, and identify the
phase shift $\Delta \varphi = 2\pi f \tau$. Comment on the sign
convention (lead versus lag).

\paragraph{Worked solution.}
\textbf{Set-up.} A pure time delay $\tau$ applied to a sinusoidal
signal produces a copy whose argument has been shifted by
$2\pi f \tau$. The relation between time delay and phase shift is
linear in frequency and is the working dictionary between
oscilloscope time-domain measurements and frequency-domain phase
diagrams.

\textbf{Given.} $x(t) = A \cos(2\pi f t)$ and $y(t) = x(t - \tau)
= A \cos(2\pi f (t - \tau))$. Compute the phase shift $\Delta
\varphi$ and identify its sign.

\textbf{Step 1: expand the delayed-signal argument.}
$2\pi f (t - \tau) = 2\pi f t - 2\pi f \tau$, so
\[
y(t) = A \cos(2\pi f t - 2\pi f \tau).
\]

\textbf{Step 2: read off the phase shift.} At any time $t$, the
instantaneous phase of $y$ is $\varphi_{y}(t) = 2\pi f t - 2\pi f
\tau$ and the instantaneous phase of $x$ is $\varphi_{x}(t) =
2\pi f t$. The difference is $\Delta \varphi = \varphi_{y} -
\varphi_{x} = -2\pi f \tau$.

\textbf{Step 3: assign the sign convention.} A positive $\tau$
delays $y$ relative to $x$; the phase of $y$ at any reference
time is smaller than the phase of $x$ at the same time. By
convention, $y$ \engterm{lags} $x$ by $|\Delta \varphi| = 2\pi
f \tau$. A negative $\tau$ (an advance) reverses the sign and
produces a lead.

\textbf{Sanity check.} A $\tau = 1/(4 f)$ delay (one quarter
period) gives $\Delta \varphi = -\pi/2$, the canonical $90^\circ$
lag of $\cos$ versus $\sin$. A full-period delay $\tau = 1/f$
gives $\Delta \varphi = -2\pi$, equivalent to no shift at all,
consistent with the periodic identity $\cos(\theta - 2\pi) =
\cos(\theta)$.

\textbf{Common error.} Conflating the sign convention for delay
with the sign convention for lead/lag. The mathematical statement
``phase shift of $-2\pi f \tau$'' is unambiguous; the operational
description ``$y$ lags $x$ by $2\pi f \tau$'' is the working
phrasing and uses the absolute value. The two never disagree, but
sliding between conventions inside a single derivation produces
sign errors. The working discipline is to declare which convention
the derivation uses at the start and to stick with it.
\end{exercise}

\begin{exercise}
The Nyquist condition asserts that a sinusoid of frequency $f$
cannot be uniquely identified from samples taken at rate
$f_{s} \leq 2f$. By writing the sampled values
$x_{n} = A \cos(2\pi f n T_{s})$ at $f = f_{s}/2 + \delta$ for
small $\delta$, show that the same sample sequence can be produced
by a sinusoid at frequency $f_{s}/2 - \delta$ with appropriate
amplitude and phase. State the alias relation that this
demonstrates.

\paragraph{Solution sketch.}
With $T_{s} = 1/f_{s}$ and $f = f_{s}/2 + \delta$,
\[
x_{n} = A \cos(2\pi (f_{s}/2 + \delta) n / f_{s})
       = A \cos(\pi n + 2\pi \delta n / f_{s}).
\]
The $\pi n$ term contributes a factor $(-1)^{n}$. Using
$\cos(\pi n + \theta) = (-1)^{n} \cos\theta$,
\[
x_{n} = (-1)^{n} A \cos(2\pi \delta n T_{s}).
\]
The alternating sign $(-1)^{n}$ is itself the sample sequence of a
$f_{s}/2$ component; combined with the slowly-varying
$\cos(2\pi \delta n T_{s})$, the product is indistinguishable from
the sample sequence of $A \cos(2\pi (f_{s}/2 - \delta) n T_{s})$
(up to amplitude and phase) for the same set of sample times. The
alias relation: a frequency $f$ above $f_{s}/2$ folds to
$f_{\text{alias}} = |f - n f_{s}|$ for the integer $n$ that puts
the result in $[0, f_{s}/2]$. The result is the formal version of
the heuristic ``samples cannot distinguish'' picture from
\S 6.4.1.
\end{exercise}

\begin{exercise}
For a quantisation step $q$ uniformly distributed on
$[-q/2, q/2]$, show that the standard deviation of the quantisation
error is $q/\sqrt{12}$. Hint: the variance of a uniform
distribution on $[-a, a]$ is $a^{2}/3$. (The general derivation
using continuous probability density functions appears in Volume II
once the necessary calculus has been introduced; the result here
follows from the variance of a uniform distribution alone.)

\paragraph{Worked solution.}
\textbf{Set-up.} An ADC quantisation step $q$ rounds a continuous
input to the nearest of a discrete set of codes. The rounding
error is the difference between the true input and the quantised
output; if the input distribution is fine enough relative to $q$,
the error is well approximated as uniformly distributed on
$[-q/2, q/2]$. The variance of a uniform random variable on
$[-a, a]$ is $a^{2}/3$ (a standard textbook result, taken as a
given here).

\textbf{Given.} Quantisation error uniformly distributed on
$[-q/2, q/2]$. Hint: $\Var[U[-a, a]] = a^{2}/3$.

\textbf{Step 1: identify the variable.} The quantisation error is
$\epsilon \sim U[-q/2, q/2]$, so $a = q/2$ in the hint.

\textbf{Step 2: substitute and compute the variance.}
$\Var(\epsilon) = (q/2)^{2}/3 = q^{2}/12$.

\textbf{Step 3: take the square root.} The standard deviation is
$\sigma_{\epsilon} = \sqrt{q^{2}/12} = q/\sqrt{12} \approx 0.289
\, q$.

\textbf{Sanity check.} The two extreme values of $\epsilon$ are
$\pm q/2$. A standard deviation of $q/\sqrt{12} \approx 0.29 q$
sits well inside the half-range, consistent with the uniform
distribution's standard deviation being below the half-width but
within a factor of two of it. A second sanity check: for a $1
\,\text{LSB}$ ADC step, the working rms quantisation error is
about $0.29 \,\text{LSB}$, the figure that appears in working
ADC datasheets as the intrinsic noise floor.

\textbf{Common error.} Writing $\sigma_{\epsilon} = q/2$, the
half-range, as if it were the standard deviation. The half-range
is the worst-case error, not the rms. Adding the worst-case error
to other uncertainty terms in quadrature over-counts the
quantisation noise by a factor of $\sqrt{12}/2 \approx 1.7$ and
inflates the combined uncertainty. The working rule is
$q/\sqrt{12}$ for combining in quadrature; the half-range $q/2$
is the bound that the engineer cites when stating a worst-case
error guarantee.
\end{exercise}

\begin{exercise}
A unit square wave with amplitude $A$ can be approximated by
\[
x_{3}(t)=\frac{4A}{\pi}\left[\sin(2\pi f_{0} t)+\tfrac{1}{3}\sin(6\pi f_{0} t)+\tfrac{1}{5}\sin(10\pi f_{0} t)\right].
\]
Evaluate $x_{3}(t)$ at $t = 1/(4 f_{0})$ and compare it with $A$.
Explain why the three-term approximation overshoots the plateau
rather than landing within $5\,\%$ of $A$. (No calculus required;
this is direct evaluation. The persistent overshoot near the step
is the Gibbs phenomenon, taken up in Volume II.)

\paragraph{Solution sketch.}
At $t = 1/(4 f_{0})$ the three sines take values
$\sin(\pi/2) = 1$, $\sin(3\pi/2) = -1$, and $\sin(5\pi/2) = 1$,
so the bracket equals $1 - 1/3 + 1/5 = 13/15$. Then
\[
x_{3}\bigl(1/(4 f_{0})\bigr) = \frac{4A}{\pi} \cdot \frac{13}{15}
   \approx 1.103 \,A.
\]
The three-term approximation overshoots the unit plateau by about
$10\,\%$. Adding more terms changes the overshoot's location but
not its magnitude: the limiting overshoot of the truncated
Fourier series of a square wave is approximately $1.179$ of the
target amplitude (about $17.9\,\%$). The persistent overshoot at
discontinuities is the Gibbs phenomenon; the formal derivation
needs the integral form of the Fourier coefficients and is in
Volume II.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the timing precision a GPS receiver must achieve for the
position solution to have $10 \,\si{\meter}$ horizontal accuracy.
State the assumption that converts a pseudorange uncertainty into a
position uncertainty. Comment on whether open-sky smartphone
horizontal accuracy at the few-metre level is consistent with
timing precision at the few-nanosecond level.

\paragraph{Worked solution.}
\textbf{Set-up.} A GPS position solution combines four
pseudoranges (or more) through a least-squares fit; the
horizontal position uncertainty $\sigma_{H}$ relates to the
pseudorange uncertainty $\sigma_{\rho}$ through
$\sigma_{H} = \mathrm{HDOP} \times \sigma_{\rho}$, where HDOP is
the horizontal dilution of precision set by satellite geometry.
The pseudorange uncertainty maps directly to receiver-clock
timing precision through $\sigma_{\rho} = c \sigma_{t}$.

\textbf{Given.} Horizontal target $\sigma_{H} = 10 \,\si{\meter}$;
working approximation HDOP $\approx 1$ (open-sky); conversion
$1 \,\si{\nano\second} \Leftrightarrow 30 \,\si{\centi\meter}$.

\textbf{Step 1: convert the position target into a pseudorange
target.} With HDOP $\approx 1$, $\sigma_{\rho} \approx \sigma_{H}
= 10 \,\si{\meter}$.

\textbf{Step 2: convert the pseudorange target into a timing
target.} $\sigma_{t} = \sigma_{\rho}/c = 10 \,\si{\meter}/
(3 \times 10^{8} \,\si{\meter\per\second}) \approx 33 \,\si{\nano
\second}$.

\textbf{Step 3: trace the consumer receiver's chain.} The L1 C/A
code chip rate is $1.023 \,\si{\mega\hertz}$, giving a chip
duration of $1 \,\si{\micro\second}$ and a chip length of $\sim
300 \,\si{\meter}$. Sub-chip timing through code-tracking-loop
correlation reaches $0.01$ to $0.1$ chip, or $3$ to $30 \,\si
{\meter}$ on the pseudorange. The carrier-phase observable closes
on millimetre-class precision but requires phase-ambiguity
resolution that consumer receivers typically do not perform.

\textbf{Sanity check.} A smartphone GPS receiver reporting
$\sim 5 \,\si{\meter}$ open-sky horizontal accuracy is consistent
with the back-of-envelope: HDOP $\approx 1.5$, $\sigma_{\rho}
\approx 3 \,\si{\meter}$, $\sigma_{t} \approx 10 \,\si{\nano
\second}$. The receiver's own oscillator (a quartz TCXO with
$10^{-6}$ fractional accuracy) cannot hold $10 \,\si{\nano\second}$
in absolute terms; the four-pseudorange solution estimates and
removes the receiver-clock bias every epoch.

\textbf{Common error.} Forgetting the HDOP factor. Under degraded
geometry (e.g., HDOP $= 5$ in an urban canyon), the same timing
precision yields $50 \,\si{\meter}$ horizontal uncertainty, not
$10 \,\si{\meter}$. The receiver's reported accuracy depends on
both the timing precision and the geometry; quoting a position
accuracy without the prevailing HDOP is incomplete.
\end{exercise}

\begin{exercise}
Estimate the relative phase shift between two AC waveforms separated
by $1 \,\si{\milli\second}$ at $50 \,\si{\hertz}$. Compare the
result to the phase resolution that a smartphone audio recording at
$44{,}100 \,\si{\hertz}$ can support, given one sample period is
$22.7 \,\si{\micro\second}$.

\paragraph{Solution sketch.}
$\Delta \varphi = 2\pi f \tau = 2\pi \times 50 \times 10^{-3} =
0.10\pi \,\si{\radian} \approx 18^{\circ}$. The phase resolution
of the recording is $2\pi \times 50 \times 22.7 \times 10^{-6}
\approx 0.0071 \,\si{\radian} \approx 0.41^{\circ}$ per sample
period at $50 \,\si{\hertz}$, so the $18^{\circ}$ shift is
resolved with roughly forty samples of margin. With sub-sample
zero-crossing interpolation the resolution improves further. The
practical limit on phase measurement with a consumer audio chain
is rarely the sample rate; it is the front-end group delay
variation between channels and the sample clock's fractional
frequency stability.
\end{exercise}

\begin{exercise}
A recorder's sample clock has fractional frequency error $10^{-5}$.
Estimate the apparent frequency error this produces when measuring
a $50 \,\si{\hertz}$ signal, and compute the accumulated timing
error over 24 hours. State what additional calibration would be
needed before claiming $1 \,\si{\milli\hertz}$ standard
uncertainty in the project record.

\paragraph{Worked solution.}
\textbf{Set-up.} A recorder samples a periodic signal at rate
$f_{s}$ generated by an on-board oscillator. If the on-board
oscillator's fractional frequency error is $\Delta f/f$, every
measured frequency is multiplied by the same fractional error: a
true $50 \,\si{\hertz}$ signal at the input appears as
$50 \times (1 + \Delta f/f)$ in the digital record. The recorder
cannot detect this bias without an external reference.

\textbf{Given.} $\Delta f/f = 10^{-5}$ on the sample clock; signal
frequency $50 \,\si{\hertz}$; recording duration $24 \,\si{\hour}
= 86{,}400 \,\si{\second}$; target standard uncertainty $1 \,\si
{\milli\hertz}$ on the reported frequency.

\textbf{Step 1: compute the apparent frequency offset.}
$\Delta f_{\text{apparent}} = 50 \times 10^{-5} = 5 \times 10^{-4}
\,\si{\hertz} = 0.5 \,\si{\milli\hertz}$. Half a millihertz is
already half the project's target uncertainty.

\textbf{Step 2: compute the cumulative timing drift.} Over $24
\,\si{\hour}$ the clock accumulates $86{,}400 \times 10^{-5} =
0.864 \,\si{\second}$ of timing error. A diurnal log uncorrected
for this drift would appear to slip by nearly a second across the
day.

\textbf{Step 3: identify the calibration paths.} Three working
references close the gap. (a) Network Time Protocol: a
laptop disciplined by NTP typically holds $\Delta f/f \approx
10^{-7}$ after a few minutes, well below the $0.05 \,\si{\milli
\hertz}$ residual on a $50 \,\si{\hertz}$ signal. (b) GPS-
disciplined oscillator: $\Delta f/f \approx 10^{-10}$, with a
residual at the microhertz level. (c) Calibration coefficient
derived from a long recording of a known reference signal
(typically a primary or secondary frequency standard); the
calibration coefficient can be applied in post-processing.

\textbf{Sanity check.} The fractional figure $10^{-5}$ is typical
of an uncalibrated PC sound-card oscillator; a TCXO improves to
$10^{-7}$; a GPSDO reaches $10^{-10}$. The mains-frequency project
target of $1 \,\si{\milli\hertz}$ on $50 \,\si{\hertz}$ is
equivalent to $\Delta f/f = 2 \times 10^{-5}$, which a TCXO
exceeds by two orders of magnitude. A naive PC sound-card
recording at $10^{-5}$ accuracy already eats half the budget; the
GPSDO is the safe choice for a 24-hour recording.

\textbf{Common error.} Confusing sample-clock accuracy with
sample-clock stability. The Allan deviation at a single
averaging time measures stability over that interval; the
fractional accuracy measures the mean offset from the nominal
rate. A clock with poor accuracy but excellent stability can be
calibrated out once; a clock with good accuracy but poor
stability cannot. The 24-hour mains-frequency project requires
both terms to be small, and the GPSDO is the working choice for
exactly this reason.
\end{exercise}

\begin{exercise}
Estimate the bandwidth occupied by an audio signal whose spectral
content extends from $20 \,\si{\hertz}$ to $20 \,\si{\kilo\hertz}$,
sampled at $44{,}100 \,\si{\hertz}$. State the margin between the
audio bandwidth and the Nyquist frequency, and identify the role
of the anti-aliasing filter in the recording chain.

\paragraph{Worked solution.}
\textbf{Set-up.} A digital audio recording captures an audio
signal whose spectral content extends from a low-frequency
cutoff $f_{L}$ to a high-frequency cutoff $f_{H}$, sampled at
$f_{s}$. The Nyquist frequency $f_{s}/2$ must exceed $f_{H}$
with enough margin to allow the anti-aliasing filter's
transition band; the difference $f_{s}/2 - f_{H}$ is the
filter designer's working budget.

\textbf{Given.} Audio band $f_{L} = 20 \,\si{\hertz}$ to $f_{H} =
20 \,\si{\kilo\hertz}$; sample rate $f_{s} = 44{,}100 \,\si
{\hertz}$ (the CD standard).

\textbf{Step 1: compute the audio bandwidth.}
$B = f_{H} - f_{L} = 20{,}000 - 20 = 19{,}980 \,\si{\hertz}
\approx 20 \,\si{\kilo\hertz}$.

\textbf{Step 2: compute the Nyquist frequency.}
$f_{N} = f_{s}/2 = 22{,}050 \,\si{\hertz}$.

\textbf{Step 3: compute the anti-alias transition margin.}
$\Delta f = f_{N} - f_{H} = 22{,}050 - 20{,}000 = 2{,}050 \,\si
{\hertz}$, about $10\,\%$ of the audio bandwidth. The anti-alias
filter must pass the audio band with acceptable amplitude and
phase response up to $f_{H}$ and reach its stopband attenuation
(typically $80$ to $100 \,\si{\decibel}$) by $f_{N}$.

\textbf{Step 4: derive the implied filter order.} A
$4\,\text{th}$-order Butterworth filter with $-3 \,\si{\decibel}$
corner at $20 \,\si{\kilo\hertz}$ provides $24 \,\si{\decibel}$
per octave of rolloff. The fractional octave between $20$ and
$22 \,\si{\kilo\hertz}$ is $\log_{2}(22/20) \approx 0.137$
octaves, giving $24 \times 0.137 \approx 3.3 \,\si{\decibel}$ of
attenuation by Nyquist, well below the $80 \,\si{\decibel}$
target. A steeper filter (elliptic, $8\,\text{th}$ order) or
oversampling with a lower analog-front-end corner is the
production solution; many modern audio interfaces use
delta-sigma modulators at $\geq 5 \,\text{MHz}$ and digital
decimation to relax the analog requirements.

\textbf{Sanity check.} The CD standard of $44.1 \,\si{\kilo
\hertz}$ was chosen for exactly this margin: it accommodates the
$20 \,\si{\kilo\hertz}$ working upper bound of audible content
with a $10\,\%$ transition band that 1980s-era analog filter
technology could realise without intruding on the audio band.
The professional audio standard of $48 \,\si{\kilo\hertz}$
provides a slightly more comfortable $4 \,\si{\kilo\hertz}$
transition band.

\textbf{Common error.} Quoting the Nyquist frequency as the
hard upper bound of the recordable band. The Nyquist frequency
is the unaliased ceiling; the working upper bound sits at $f_{H}
< f_{N}$ to leave room for the anti-alias filter's transition.
The reader who records a $22 \,\si{\kilo\hertz}$ test tone at
$44.1 \,\si{\kilo\hertz}$ and discovers the tone is severely
attenuated has met the anti-alias filter, not the Nyquist
condition.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
In a language of the reader's choice, generate a $5 \,\si{\second}$
sinusoidal signal at $f = 50 \,\si{\hertz}$ sampled at $f_{s} =
1 \,\si{\kilo\hertz}$. Compute the discrete Fourier transform and
plot the magnitude spectrum. Identify the bin at which the peak
falls and the frequency it corresponds to. Repeat with the signal
shifted to $f = 50.3 \,\si{\hertz}$ and observe the spectral
leakage into adjacent bins.

\paragraph{Solution sketch.}
$N = f_{s} \times 5 = 5{,}000$ samples; bin spacing $\Delta f =
f_{s}/N = 0.2 \,\si{\hertz}$. At $f = 50 \,\si{\hertz}$ the
energy lands cleanly in bin $k = f / \Delta f = 250$. At $f =
50.3 \,\si{\hertz}$ the peak lies between bins 251 and 252; the
energy is shared across several neighbours, with sinc-shaped
sidelobes from the implicit rectangular window. A Hann window
applied before the DFT narrows the leakage at the cost of a
slightly wider main lobe. The script at
\texttt{code/fft\_noisy\_signal.py} runs a one-second variant with
parabolic peak interpolation; adapting it to the five-second case
amounts to changing the duration parameter.
\end{exercise}

\begin{exercise}
Generate a sinusoid at $f = 7{,}500 \,\si{\hertz}$ and sample it at
$f_{s} = 8{,}000 \,\si{\hertz}$. Plot the sampled values overlaid
on the original sinusoid for one second. Identify the alias
frequency at which the samples appear to oscillate. Compute the
alias frequency analytically and verify it matches the visual
result.

\paragraph{Solution sketch.}
$f_{s}/2 = 4{,}000 \,\si{\hertz}$; with $n = 1$ the alias is
$|7{,}500 - 8{,}000| = 500 \,\si{\hertz}$. The sampled record
shows a clean $500 \,\si{\hertz}$ sinusoid. The script at
\texttt{code/aliasing\_demo.py} runs this exact scenario and
verifies bit-equality of the sampled sequence with the alias
sinusoid (up to a global sign).
\end{exercise}

\begin{exercise}
Simulate a 100-hour Patriot-style integer counter incrementing at
$10 \,\si{\hertz}$, with each increment converted to seconds by
multiplying by a $24$-bit truncated binary representation of $1/10$.
Plot the cumulative error versus elapsed time. Verify that the
error after $100 \,\si{\hour}$ is approximately $0.34 \,\si
{\second}$. Comment on what changes if the truncation is performed
at $32$ or $64$ bits instead.

\paragraph{Solution sketch.}
The truncated $24$-bit representation of $1/10$ is $0.0999999940
\ldots$, a per-tick error of about $9.5 \times 10^{-8} \,\si
{\second}$. Over $100 \,\si{\hour} = 3.6 \times 10^{6} \,\si
{\second}$ at $10 \,\si{\hertz}$ the counter accumulates $3.6
\times 10^{7}$ ticks; the cumulative error reaches $3.6 \times
10^{7} \times 9.5 \times 10^{-8} \approx 0.34 \,\si{\second}$,
matching the GAO report. Doubling the truncation width to $32$
bits drops the per-tick error by a factor of $2^{8} = 256$ and
the $100 \,\si{\hour}$ cumulative error to about $1.3 \,\si
{\milli\second}$; $64$ bits reduces it below picosecond levels
over the same interval. A double-precision IEEE 754 representation
of $0.1$ has roughly $52$-bit mantissa precision and a per-tick
error around $10^{-17} \,\si{\second}$, negligible over operating
lives much longer than any realistic mission. The script at
\texttt{code/patriot\_drift.py} runs all four cases (16, 24, 32,
64 bits) and tabulates the results.
\end{exercise}

\begin{exercise}
Simulate a noisy near-$50 \,\si{\hertz}$ signal at sample rate
$10 \,\si{\kilo\hertz}$ with additive white Gaussian noise at
signal-to-noise ratio of $40 \,\si{\decibel}$. Estimate the
frequency by zero-crossing detection (with linear interpolation
between samples) and by FFT peak detection over windows of $1$ and
$60 \,\si{\second}$. Compare the standard deviations of the three
estimators across $1{,}000$ trials.

\paragraph{Solution sketch.}
At $40 \,\si{\decibel}$ SNR the noise rms is about $1\,\%$ of the
signal amplitude. The Cramer-Rao lower bound for the frequency
of a sinusoid in white Gaussian noise scales as $T^{-3/2}$ in the
record length $T$. The three expected standard deviations are
approximately: zero-crossing with linear interpolation over $60
\,\si{\second}$, around $10^{-4} \,\si{\hertz}$; FFT peak (raw bin)
over $1 \,\si{\second}$ at bin spacing $1 \,\si{\hertz}$, around
$10^{-1} \,\si{\hertz}$ without interpolation, around $10^{-3} \,
\si{\hertz}$ with parabolic peak interpolation; FFT peak with
parabolic interpolation over $60 \,\si{\second}$ at bin spacing
$1/60 \,\si{\hertz}$, around $10^{-5}$ to $10^{-4} \,\si{\hertz}$.
The ranking is sensitive to noise colour, window function, and
whether the estimator is unbiased; the reader's empirical numbers
are the answer. The scripts at \texttt{code/mains\_zero\_crossing.py}
and \texttt{code/fft\_noisy\_signal.py} implement the first and
second estimators respectively; the trial loop is left to the
reader.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a measurement system to record the grid frequency at a
substation, current as of 2026, at one-second resolution with
combined standard uncertainty of $1 \,\si{\milli\hertz}$ per
sample. State the sensor (or coupling), the sample rate, the
analog conditioning, the sample-clock requirement, the data
storage, and the calibration procedure. Identify the dominant
expected uncertainty source.

\paragraph{Discussion.}
A workable answer specifies an isolated potential-transformer
coupling on a substation bus, an anti-aliased low-pass front end
with corner well below the Nyquist frequency, a sample rate of
$10 \,\si{\kilo\hertz}$ or higher (well above the few-hertz
signal bandwidth, with margin for any harmonic content of
interest), a sample clock disciplined to UTC through a GPSDO with
fractional stability $\leq 10^{-10}$, on-board storage of derived
frequency estimates rather than raw waveforms, and an annual
substitution calibration against a known reference frequency
source. A defensible response identifies the sample-clock
stability as the dominant uncertainty source if it is not GPS-
disciplined, the estimator's algorithm bias (zero-crossing vs.
fitted sinusoid vs.\ FFT peak) if it is. A weaker response
specifies only the sample rate and ignores the clock; an
acceptable response addresses all six items.
\end{exercise}

\begin{exercise}
Design a digital audio recording chain for spoken voice with a
required intelligibility-grade signal-to-noise ratio of
$30 \,\si{\decibel}$ over the band $300$ to $3{,}400 \,\si{\hertz}$.
Specify the sample rate, the bit depth, the anti-aliasing filter
order and corner frequency, the pre-amplifier gain, and the
storage format. Justify each choice against the application.

\paragraph{Discussion.}
The G.711 telephony defaults are a defensible starting point:
sample rate $8 \,\si{\kilo\hertz}$ (Nyquist $4 \,\si{\kilo\hertz}$,
covering the $3.4 \,\si{\kilo\hertz}$ upper edge with margin);
bit depth $8$ bits with $\mu$-law companding or $16$ bits linear,
the latter giving comfortable SNR head-room above the
$30 \,\si{\decibel}$ requirement; anti-aliasing filter of $4\,
\text{th}$ to $6\,\text{th}$ order with corner near $3.6 \,\si
{\kilo\hertz}$ and stopband attenuation by $4 \,\si{\kilo\hertz}$;
pre-amplifier gain set so that the expected speech-signal peaks
exercise most of the $\pm V_{\text{range}}/2$ range without
clipping. Storage format depends on downstream use: linear PCM
WAV for laboratory use, $\mu$-law or A-law for telephony
compatibility, Opus or AAC for size-constrained transport. The
exercise lets the reader defend trade-offs explicitly; an
acceptable answer matches each choice to a numerical requirement.
\end{exercise}

\begin{exercise}
Design a clock synchronisation scheme for a network of three
distributed sensors that must produce timestamps consistent to
within $1 \,\si{\milli\second}$ across a working day. State the
synchronisation reference (GPS, network time service, hardware
clock distribution), the resynchronisation interval, the local
oscillator class, and the uncertainty budget across the network.

\paragraph{Discussion.}
A working-day timing budget of $1 \,\si{\milli\second}$ allows
many viable answers. A defensible scheme uses NTP discipline
against a stratum-1 reference over the local network with a
quartz TCXO local oscillator (fractional accuracy $\sim 10^{-7}$
to $10^{-8}$); the resynchronisation interval can be tens of
seconds to minutes, leaving substantial margin against the $1 \,
\si{\milli\second}$ budget. A more disciplined scheme uses
PTP (IEEE 1588) over a network that supports hardware
timestamping at the switches; this routinely reaches sub-
microsecond cross-network consistency. A GPSDO at each node
gives microsecond-class consistency end to end, at higher
hardware cost. The uncertainty budget aggregates: per-node
oscillator drift over the resynchronisation interval, network
asymmetry (the dominant NTP error term), and clock-step
discontinuities. A weaker response specifies only the reference;
an acceptable response addresses all four items and constructs
the budget explicitly.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A $50 \,\si{\hertz}$ mains signal sampled at $40 \,\si{\hertz}$ is
displayed on a software oscilloscope as a clean $10 \,\si{\hertz}$
sinusoid. State the cause and propose the next measurement that
would distinguish the alias from a genuine $10 \,\si{\hertz}$
signal.

\paragraph{Solution sketch.}
$f_{s}/2 = 20 \,\si{\hertz}$ and $f = 50 \,\si{\hertz}$; with $n
= 1$, $f_{\text{alias}} = |50 - 40| = 10 \,\si{\hertz}$. The
displayed $10 \,\si{\hertz}$ is the alias of the mains. The
diagnostic test is to change the sample rate, say to $100 \,\si
{\hertz}$ or $1 \,\si{\kilo\hertz}$. A genuine $10 \,\si{\hertz}$
signal still appears at $10 \,\si{\hertz}$; the alias shifts to a
new (correct) frequency above the new Nyquist or appears at $50
\,\si{\hertz}$ if the new rate exceeds $100 \,\si{\hertz}$.
Confirming an anti-aliasing filter at the front end is the second
diagnostic.
\end{exercise}

\begin{exercise}
A laboratory data-acquisition system samples a thermocouple-
amplifier output at $1 \,\si{\hertz}$. The recorded signal shows
periodic excursions whose amplitude is much larger than the
thermal noise floor and whose frequency is exactly $1 \,\si
{\hertz}$ in the displayed signal. State three plausible mechanisms
(mains-frequency interference aliased to DC, ground-loop pickup at
the sample rate, thermocouple cold-junction modulation, software
sampling phase coupled to a periodic process) and propose the
next measurement that would distinguish them.

\paragraph{Solution sketch.}
A $1 \,\si{\hertz}$ excursion in a $1 \,\si{\hertz}$-sampled
record can be many things; that the displayed frequency is
exactly the sample frequency itself is the diagnostic clue. A
mains-frequency interference at $50 \,\si{\hertz}$ sampled at $1
\,\si{\hertz}$ aliases to $|50 - 50 \times 1| = 0 \,\si{\hertz}$
(DC) for an exact integer ratio, but jitter on the sample clock
spreads that DC alias into a slow modulation. Any periodic
process locked to the sample-clock phase (a software task that
runs every second and disturbs the front end, a switching power
supply with a $1 \,\si{\hertz}$ envelope, a cooling fan whose
rotation rate happens to be locked to a $1 \,\si{\hertz}$
controller) appears at the sample rate exactly. The diagnostic is
to sample at a non-commensurate rate (say $1.1 \,\si{\hertz}$ or
$0.7 \,\si{\hertz}$); a genuine periodic process at a fixed
frequency shifts in the displayed record, while a sample-rate-
locked artefact stays at the new sample rate.
\end{exercise}

\begin{exercise}
A GPS-disciplined oscillator at a substation reports a sudden
$50 \,\si{\nano\second}$ jump in its time output. State three
plausible mechanisms (a satellite outage with handover to a worse
geometry, a multipath excursion, a receiver firmware update, a
local oscillator step) and propose the diagnostic test sequence
that would distinguish them.

\paragraph{Worked solution.}
\textbf{Set-up.} A GPSDO at a substation reports a sudden $50 \,\si
{\nano\second}$ step in its disciplined output. The step is large
enough to violate the substation's microsecond timing budget if it
persists. The four candidates (satellite handover, multipath
excursion, firmware update, local oscillator step) each have a
distinct signature in the receiver's auxiliary records; the
diagnostic strategy is to query each record in increasing order
of operator effort.

\textbf{Given.} GPSDO time-output step of $50 \,\si{\nano\second}$
(corresponding to about $15 \,\si{\meter}$ on a single
pseudorange); substation context.

\textbf{Step 1: query the receiver event log for firmware updates.}
A firmware update typically resets the holdover state and may
introduce a single discontinuity in the disciplined output. The
event log is the cheapest record to query (read-only, near-real-
time, no signal processing required). If the log shows an update
at or near the step time, the firmware update is the cause.

\textbf{Step 2: inspect the satellite-tracking record.} A
satellite handover (a tracked satellite drops below the elevation
mask and is replaced by another) shifts the geometric dilution
of precision. The signature is a step at the handover boundary,
visible in the receiver's satellite-tracking log alongside a
change in the satellite identifier set. The post-fit residuals
on the dropped satellite typically grow before the handover.

\textbf{Step 3: examine the carrier-phase multipath estimator.} A
multipath excursion (a reflected signal arrives slightly later
than the direct signal) is transient over minutes and is visible
in the carrier-phase multipath estimator as a localised
excursion that decays back to baseline. If the step persists
beyond several minutes, multipath is not the cause.

\textbf{Step 4: examine the disciplining loop's steering signal.}
A local oscillator step (a sudden frequency change on the
disciplined OCXO, possibly from a temperature shock or
vibration) appears in the disciplining loop's steering voltage
as a jump in the integrator output. The loop chases the
oscillator step over its time-constant (typically tens of
seconds for a substation-grade GPSDO), leaving a transient on
the output.

\textbf{Step 5: cross-compare against a second GPSDO.} A critical
timing installation carries at least two GPSDOs for exactly this
kind of diagnosis. If the second GPSDO does not show the step,
the cause is local to the first unit (firmware, oscillator,
multipath at one antenna); if both step together, the cause is
common (satellite handover, ionospheric event, broadcast clock
error).

\textbf{Sanity check.} The four candidates produce signatures
that are mutually exclusive at the auxiliary-record level: only
one of the four logs typically shows the step. If two logs
agree (e.g., the firmware update happened at the same time as a
satellite handover), the engineer should suspect the firmware
update first (it is the more common precipitating event).

\textbf{Common error.} Treating any step in the GPSDO output as
a satellite-side problem. The four candidates partition between
receiver-side (firmware, oscillator) and signal-side (handover,
multipath) causes. The receiver-side causes are more common in
operations and easier to fix; the discipline is to query the
event log first, before reaching for the satellite-tracking
record.
\end{exercise}

\begin{exercise}
A digital recording of a $50 \,\si{\hertz}$ mains signal shows a
strong peak at $50 \,\si{\hertz}$ and smaller peaks at $150$,
$250$, and $350 \,\si{\hertz}$. State whether the smaller peaks
are aliasing artefacts, harmonics of a non-sinusoidal mains
waveform, or interference from another source, and propose the
measurement that distinguishes them.

\paragraph{Solution sketch.}
The smaller peaks at $150$, $250$, and $350 \,\si{\hertz}$ are
the third, fifth, and seventh harmonics of the $50 \,\si{\hertz}$
fundamental, the signature of a non-sinusoidal waveform with
odd-harmonic symmetry (the same pattern as a square wave). The
mains waveform in many synchronous areas carries odd-harmonic
distortion at the percent level, driven by non-linear loads
(switching power supplies, fluorescent lamps, motor drives). The
diagnostic to confirm: change the sample rate; the harmonics
remain at the same frequencies (in hertz), while aliases would
shift to new locations. To distinguish harmonics of the mains
from harmonics generated locally, observe the same recording at
a different location on the same grid; mains-borne harmonics are
common to the location, locally generated ones are not.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
The Patriot case is canonical for accumulated arithmetic error in a
long-running clock. Argue, in 200 to 300 words, what additional
verification or operational discipline would have detected the
mechanism before the Dhahran incident, citing the GAO report
\cite{acc:gao-patriot-1992}.

\paragraph{Worked solution.}
\textbf{Set-up.} The Patriot's clock-error mechanism (a fixed-point
representation of $0.1$ that drifts at $9.5 \times 10^{-8}\,\si
{\second}$ per tick) is detectable, in principle, by any of
several engineering practices that the system did not run. The
question is which practice, applied during development or
operations, would have surfaced the error before Dhahran.

\textbf{Given.} The cumulative-drift trajectory ($0.34 \,\si
{\second}$ at $100 \,\si{\hour}$, growing linearly in time) and
the institutional context (the patch arrived the day after the
incident).

\textbf{Step 1: continuous-operation soak test.} A continuous-
operation test at or beyond the longest expected mission duration,
with the cumulative-error metric instrumented against an external
reference (a GPS-disciplined oscillator was widely available by
1991), would have logged the linear drift directly. The test cost
is modest: a single battery in test configuration, $100 \,\si
{\hour}$ of operating time, an external timing reference, and
post-processing. The discipline would surface the drift signature
once the operating envelope exceeded the qualifier-test envelope.

\textbf{Step 2: static analysis of the number-representation
chain.} The fixed-point representation of $0.1$ in binary is
non-terminating; a static analyser that flags exact-decimal-to-
binary conversions and bounds their per-step error would have
flagged the conversion as a known precision risk. The tools for
this analysis existed in 1991 (Ada's strong typing, Fortran
INTENT, hand-derivation of error bounds); the discipline did not.

\textbf{Step 3: runtime invariant check.} A runtime check that
compares two independent time-conversion paths (an integer-only
chain against the floating-point chain) and triggers on
disagreement past a threshold would have flagged the cumulative
mismatch within minutes of system startup. The implementation is
small (a few hundred bytes of code, a millisecond per check) and
the alarm is actionable.

\textbf{Step 4: operational discipline tying mission length to
allowed drift.} The deployment instruction permitted continuous
operation past the qualifier-test envelope. An operational rule
that bounded the maximum continuous operating time as a function
of the design's per-tick error and the system's tolerance to
that error would have triggered a documentation review when the
Dhahran battery's operating time exceeded the bound.

\textbf{Step 5: synchronised patch and operational guidance.}
The GAO report identifies the patch arriving the day after the
incident as the central institutional failure
\cite{acc:gao-patriot-1992}. A joint-distribution discipline that
ties software changes to operational guidance, so that operators
do not extend deployment envelopes past unpatched software
versions, would have prevented the mismatch in time.

\textbf{Sanity check.} The five disciplines (soak test, static
analysis, runtime invariant, operational bound,
patch-guidance joint distribution) decompose the problem along
the lifecycle stages: development, verification, operations,
deployment, support. A complete answer flags at least one
discipline at each stage; the deepest answers recognise that the
mechanism is intrinsic to fixed-point arithmetic with
non-power-of-two divisors and generalises to inertial navigation
(roll-up integration errors), financial accounting (penny
truncation), and power-grid time stamping (sample-rate-locked
artefacts in PMU networks).

\textbf{Common error.} Treating the Patriot failure as a software
bug to be patched, rather than as a system-level design and
operations failure. The patch existed in time; the failure was
that the operating envelope exceeded the qualifier-test envelope
without anyone in the chain treating the mismatch as
load-bearing. The general lesson is that long-running
arithmetic-error mechanisms are detectable by engineering
discipline at every stage of the lifecycle; the operational
deployment must respect the verified envelope, and the patch
distribution must arrive synchronised with the operational
guidance.
\end{exercise}

\begin{exercise}
A power-grid phasor measurement unit (PMU) network shows a
time-of-day-correlated bias in phasor angle measurements that
appears at all PMUs in the network. Argue, in 200 to 300 words,
the failure modes that could produce a common bias across the
network and the additional measurement diversity (different
synchronisation reference, different local oscillator class,
different antenna geometry) that would close each gap.

\paragraph{Worked solution.}
\textbf{Set-up.} A common bias across an entire PMU network at the
same time of day is the signature of a shared cause; PMU
independence is broken by a common reference, a common firmware,
or a common environmental factor. The diagnostic frame mirrors
the AF447 sensor-diversity exercise from Chapter 5: redundant
channels share a hidden common cause, and the discipline of
auditing the shared causes is the same across domains.

\textbf{Given.} A power-grid PMU network shows a time-of-day-
correlated bias in phasor angle measurements that appears at
all PMUs simultaneously.

\textbf{Step 1: common GPS time-transfer chain.} All PMUs may use
GPS as the primary time reference; an ionospheric-delay
correction error, a broadcast clock or ephemeris error, or a
satellite handover correlated across PMUs shifts all receivers'
time-of-day estimates together.
\emph{Gap closer}: pair GPS-disciplined PMUs with PMUs disciplined
from a different GNSS (Galileo, BeiDou, or GLONASS) or from a
fibre-based time-transfer link (PTP over a synchronous Ethernet
or a calibrated fibre time-transfer service). The independent
chain breaks the GPS common-mode.

\textbf{Step 2: common firmware behaviour.} The PMU fleet is
typically deployed with the same firmware version, the same
calibration coefficient, the same time-zone or leap-second
handling. A shared firmware bug appears at all PMUs
simultaneously.
\emph{Gap closer}: deploy at least two PMU vendors in the
network, with diverse firmware lineages. The cost is a more
complex integration but the gain is independence from
single-firmware bugs.

\textbf{Step 3: network-wide NTP fallback.} When GPS quality
degrades, many PMUs fall back to NTP over the substation
network; the NTP server's own bias then propagates to every PMU
that fell back.
\emph{Gap closer}: ensure the NTP fallback uses a stratum-1
reference traceable to UTC, and document the fallback's
expected bias; or use a hardware-distributed timing fabric
(PTP over hardware-timestamped network) as the primary, not the
fallback.

\textbf{Step 4: time-of-day-correlated mains harmonic content.}
The mains waveform's harmonic content follows the load curve;
PMU algorithms with sensitivity to harmonic content produce a
time-of-day-correlated bias even when the timing reference is
ideal.
\emph{Gap closer}: validate the PMU algorithm against synthetic
waveforms with realistic harmonic content; adjust the windowing
or filtering to make the algorithm robust against time-of-day
harmonic variation.

\textbf{Step 5: holdover-behaviour commonality.} All PMUs use the
same local oscillator class (e.g., quartz OCXO) with the same
holdover dynamics; when the network loses GPS together (e.g.,
during a coordinated jamming event), all PMUs drift together
according to the same holdover model.
\emph{Gap closer}: pair quartz OCXO local oscillators with
chip-scale atomic local oscillators (CSACs) on a subset of PMUs
to break the holdover commonality; the CSAC-equipped PMUs
remain reliable during longer GPS outages.

\textbf{Sanity check.} The PMU case mirrors the AF447 substation
voltage-transducer case from Chapter 5 in a different domain.
The structural pattern is identical: redundant channels with
shared cause are not independent. The four-dimension diversity
framework (synchronisation reference, local oscillator class,
antenna geometry, vendor and firmware) audits the shared causes
systematically.

\textbf{Common error.} Quoting PMU network redundancy without
auditing the shared synchronisation chain. The PMU vote against
single-channel failure is necessary, not sufficient; without
diversity in synchronisation reference, local oscillator class,
and firmware lineage, the network is exposed to common-cause
events that the vote does not catch. The IEEE C37.118 standard's
microsecond accuracy requirement is met by GPSDO architectures
specifically designed for substation environments; the
discipline of auditing the common chain remains the operator's
responsibility.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer proposes to record a $40 \,\si{\kilo\hertz}$
ultrasonic signal using a $44{,}100 \,\si{\hertz}$ smartphone audio
recording. Write a one-paragraph response that takes the proposal
seriously, identifies the regime in which it could work, and
identifies the regime in which it cannot.

\paragraph{Discussion.}
A defensible response: $f = 40 \,\si{\kilo\hertz}$ is below the
Nyquist frequency of $22.05 \,\si{\kilo\hertz}$ only if the
smartphone's input chain admits the signal in the first place,
and smartphone microphones and front-end anti-aliasing filters
are typically rolled off well below $20 \,\si{\kilo\hertz}$. If
the smartphone's analog chain does pass the $40 \,\si{\kilo
\hertz}$ component (uncommon), the sampler will alias it to
$|40 - 44.1| = 4.1 \,\si{\kilo\hertz}$, which is recoverable only
if no real signal lives at $4.1 \,\si{\kilo\hertz}$. The proposal
therefore works only as a coarse presence detector under
restrictive assumptions; it cannot serve as a quantitative
ultrasonic measurement. The professional recommendation is a
sample rate of $\geq 96 \,\si{\kilo\hertz}$ with an ultrasound-
capable transducer and an explicit anti-aliasing filter; many
professional audio interfaces meet this. The response takes the
junior engineer's instinct seriously (use what is to hand)
without endorsing a method that quietly fails. An acceptable
answer is one paragraph that names both the alias mechanism and
the front-end limitation.
\end{exercise}

\begin{exercise}
Argue, in 200 to 300 words, when GPS is an acceptable time
reference for an engineering system whose timing requirement is at
the microsecond level. Take a position for one class of system
(power-grid synchrophasors, financial trading, telecommunications
backhaul, scientific instrumentation) where GPS timing is
acceptable with proper antenna, holdover, and monitoring design,
and one class where GPS alone is insufficient because jamming,
spoofing, indoor operation, certification, or holdover
requirements dominate.

\paragraph{Discussion.}
A defensible position favours GPS-with-discipline for power-grid
synchrophasors: outdoor rooftop antennas with sky view, redundant
GPSDOs, a holdover oscillator (OCXO or chip-scale atomic) that
maintains the microsecond budget across the expected loss-of-
signal interval, an interference and integrity monitor, and a
documented alternate timing plan. The IEEE C37.118 standard sets
the microsecond timing requirement for which GPSDO architectures
were designed, and the antenna environment at a substation is
typically benign. The same defensible position rejects GPS as a
sole reference for financial trading systems in dense urban
co-location facilities, where indoor operation, jamming and
spoofing exposure, regulatory traceability requirements (MiFID II
RTS 25 in Europe, equivalent rules elsewhere), and millisecond-
to-microsecond audit requirements demand a hardware-distributed
timing fabric (Precision Time Protocol over hardware-timestamped
networks, with traceable GPS or atomic upstream sources) rather
than per-server GPS. An answer that takes a clear position for
one class and against another, with the specific environmental
and regulatory constraints named, is acceptable. An answer that
collapses both into ``it depends'' or that does not address the
holdover, antenna, and interference monitoring is incomplete.
\end{exercise}
