\chapter{Phase diagrams}
\label{vol05:ch04}

\epigraph{The state of a body, however complex its composition, can be
characterized by the number of independent variables and the number of
phases coexisting in equilibrium.}{J. Willard Gibbs's phase rule
$F = C - P + 2$, given the modern engineering statement in
\textcite[ch.~9]{text:porter-easterling-phase}.}

\chapmeta{Half-life: foundational. Archetypes: stability, scaling. Exercise target: 31. Project track: physical (heat treatment and hardness testing of a small steel sample). Hazard class: moderate (small-furnace temperatures up to $900\,\degC$; quench medium handling; safety glasses and tongs required). Reader path default: standard, with sections explicitly tagged. Named cases: none required; improper-heat-treatment failures are described by mechanism. Three Mile Island (1979) is cross-referenced as a related thermal-history failure.}

\noindent
A phase diagram is a map of what arrangement of matter is
thermodynamically stable at each combination of temperature, pressure,
and composition. The chapter develops the diagrams that engineers
read off in selecting materials, planning heat treatments, controlling
solidification of castings and welds, choosing solder compositions,
and predicting how an alloy will respond to thermal excursion in
service. The chapter starts with single-component diagrams, moves to
binary and ternary diagrams, develops the lever rule for tie-line
compositions, names the eutectic and peritectic invariant reactions,
introduces time-temperature-transformation (TTT) diagrams for
non-equilibrium cooling, applies the machinery to the iron-carbon
system that underwrites every steel heat treatment, and closes with
the failure mode of incorrect or interrupted heat treatment that
produces brittle parts in service.

\section{Single-component phase diagrams}\pathtag{core}

For a pure substance, the equilibrium phase as a function of temperature
and pressure is shown on a two-dimensional $p$-$T$ diagram. The diagram
consists of regions (each one labelled with a single phase: solid,
liquid, vapour, possibly several solid polymorphs), lines that separate
the regions (each one labelled with the two-phase equilibrium that
holds along it: sublimation, melting, vaporisation, solid-solid
transitions), and points where three or more lines meet
(\engterm{triple points}, where three phases coexist).

\engterm{The water phase diagram}. The most familiar example. Solid
ice, liquid water, and water vapour occupy three regions, separated by
the ice-water melting line, the water-vapour vaporisation line, and
the ice-vapour sublimation line. The triple point of water is at $T_3
= 0.01\,\degC = 273.16\,\si{\kelvin}$ and $p_3 = 611.66\,\text{Pa}$;
the critical point, beyond which liquid and vapour are
indistinguishable, is at $T_c = 374\,\degC$ and $p_c = 22.1\,\text{MPa}$.
The melting line of water has a small negative slope ($dT/dp < 0$),
characteristic of substances that expand on freezing; ice cubes float
because the solid is less dense than the liquid. Most other materials
have positive melting-line slopes (the solid is denser than the
liquid). Higher-pressure regions of the water phase diagram contain
multiple ice polymorphs (ice II, III, V, VI, VII), distinguished by
the geometric arrangement of the hydrogen-bonded network; the
high-pressure ices are mostly outside engineering practice but appear
in planetary science and in some high-pressure laboratory contexts.

\engterm{The Clausius-Clapeyron equation}. The slope of a phase
boundary in the $p$-$T$ plane is set by the latent heat and the
volume change of the transition,
\[
\frac{dp}{dT} = \frac{\Delta H_{\text{tr}}}{T \Delta V_{\text{tr}}},
\]
with $\Delta H_{\text{tr}}$ the latent heat (positive for the higher-
entropy phase having more energy) and $\Delta V_{\text{tr}}$ the
volume change. For the vaporisation line of an ideal gas with
$\Delta V \approx V_g = RT/p$, the equation integrates to the
Clausius-Clapeyron form
\[
\ln\frac{p_2}{p_1} = -\frac{\Delta H_{\text{vap}}}{R}\left(\frac{1}{T_2}
- \frac{1}{T_1}\right),
\]
which predicts vapour pressure as a function of temperature. The
equation is the working tool for estimating boiling points at
non-standard pressure, vapour pressures in process design, and
sublimation rates in freeze-drying.

\engterm{Allotropic transitions in solids}. Many pure elements have
multiple solid phases (allotropes). The most engineering-significant
allotropic transition is iron: $\alpha$-iron (BCC, ferritic, stable
below $912\,\degC$), $\gamma$-iron (FCC, austenitic, stable from
$912\,\degC$ to $1394\,\degC$), $\delta$-iron (BCC again, stable from
$1394\,\degC$ to the melting point at $1538\,\degC$). The
$\alpha$-to-$\gamma$ transition at $912\,\degC$ is the entry point of
every steel heat treatment, because the $\gamma$-iron phase dissolves
much more carbon than the $\alpha$-iron phase, and the rapid cooling
of carbon-supersaturated $\gamma$-iron is the source of martensite.
Tin has $\alpha$-tin (grey, semiconducting, diamond-cubic structure,
stable below $13.2\,\degC$) and $\beta$-tin (white, metallic,
tetragonal, stable above $13.2\,\degC$); the $\beta$-to-$\alpha$
transition in cold-stored tin objects is called ``tin pest'' and
historically destroyed organ pipes, tin buttons, and tin solder joints
in cold climates. Carbon has graphite (the stable phase at room
temperature and pressure) and diamond (metastable at room conditions,
kinetically stable for geological times because the bond rearrangement
is hindered).

\engterm{The Gibbs phase rule}. For a system of $C$ components in
equilibrium with $P$ phases, the number of intensive degrees of
freedom is
\[
F = C - P + 2,
\]
where the $2$ accounts for the two natural intensive variables
temperature and pressure. For a single-component system ($C = 1$),
a single phase has $F = 2$ (any combination of $T$ and $p$ in the
region), two phases have $F = 1$ (the phase boundary is a line), and
three phases have $F = 0$ (the triple point). For a binary system
($C = 2$) the rule gives an extra degree of freedom (the composition),
and three-phase equilibria become lines (invariant in $T$ at constant
$p$, with specific compositions for each of the three phases). The
phase rule is the bookkeeping that limits how many phases can
coexist at a chosen condition.

\begin{archetype}[Stability]
A phase diagram is the engineering statement of which arrangement of
matter is stable at each condition. Stability here is thermodynamic
(minimum Gibbs free energy at given $T$, $p$, composition);
kinetic stability is a separate question (Chapter~2 of this volume).
The reader who confuses the two ends up surprised when an
``unstable'' diamond persists for geological times or when a
``stable'' steel surface oxidises in minutes. Phase diagrams
predict the destination; kinetics predicts the rate.
\end{archetype}

\section{Binary phase diagrams}\pathtag{core}

A binary phase diagram shows phase regions on a temperature-composition
plane (at fixed pressure, usually atmospheric). The composition axis is
typically the mass or mole fraction of one of the two components,
running from $0$ on the left (pure component A) to $1$ on the right
(pure component B). The diagram has regions of single phases (labelled
with the phase name, e.g. liquid L, solid $\alpha$, solid $\beta$),
regions of two-phase coexistence (labelled with the two phases, e.g.
$L + \alpha$, $\alpha + \beta$), and lines separating the regions.

\engterm{Isomorphous systems}. The simplest binary diagram is the
\engterm{isomorphous system}: two components that mix completely in
both the liquid and the solid phases. The diagram has only a liquid
region above the \engterm{liquidus} line, a solid region below the
\engterm{solidus} line, and a two-phase $L + \alpha$ region between
them. The copper-nickel system is the canonical example: copper melts
at $1085\,\degC$, nickel at $1455\,\degC$, and any intermediate
composition has a liquidus and a solidus between these limits, joined
at the pure-component endpoints. The two-phase region is a ``lens''
shape connecting the two melting points. The copper-nickel system
covers all alloy concentrations because the FCC lattices of copper and
nickel have nearly identical lattice constants and the two atomic
radii differ by only $2\,\%$, satisfying the Hume-Rothery rules for
extensive solid solubility.

\engterm{Eutectic systems}. Many binary systems have limited mutual
solid solubility. The simplest case is the \engterm{eutectic system},
in which the liquidus has a minimum at a particular composition (the
\engterm{eutectic composition}) and the solid below the minimum is a
two-phase mixture of two solid solutions. The lead-tin system, the
canonical solder alloy, has eutectic at composition $61.9\,\%$ tin
by mass and temperature $183\,\degC$, well below the melting points
of either pure component ($\text{Pb}$ at $327\,\degC$, $\text{Sn}$ at
$232\,\degC$). The eutectic point is invariant: three phases (liquid,
$\alpha$ solid solution rich in lead, $\beta$ solid solution rich in
tin) coexist at exactly $183\,\degC$ and $61.9\,\%$ tin. The
\engterm{eutectic reaction} on cooling is
\[
L_{\text{eut}} \to \alpha + \beta,
\]
both solids forming simultaneously from the liquid. The resulting
microstructure is a lamellar interleaving of $\alpha$ and $\beta$
phases, with a layer spacing set by the cooling rate (fine at fast
cooling, coarse at slow). Lead-tin eutectic solder was the standard
electronics-soldering material for decades; lead has been removed
from most consumer electronics under environmental regulation
(RoHS in the European Union from 2006, with global propagation), and
modern lead-free solders are tin-silver-copper eutectics or near-
eutectics at slightly higher melting temperatures ($\sim 217\,\degC$
for Sn-3.0Ag-0.5Cu).

\engterm{Peritectic systems}. A \engterm{peritectic reaction} on
cooling is
\[
L + \alpha \to \beta,
\]
where a liquid plus an existing solid phase react to form a new solid
phase at a specific temperature and composition. The reaction is the
``opposite'' of a eutectic in that the new phase forms by reaction of
a phase that previously existed with the liquid, not by direct
solidification from the liquid. Peritectic reactions are common in
iron-carbon and in copper-zinc systems. The peritectic reaction is
slow because the reaction product $\beta$ forms as a shell around
the existing $\alpha$ phase and the reaction is limited by diffusion
through this shell; non-equilibrium cooling typically leaves
unreacted $\alpha$ cores. Engineering steels that pass through a
peritectic region during solidification show characteristic core-
shell structures unless given long homogenisation heat treatments.

\engterm{Eutectoid and peritectoid reactions}. The eutectoid is the
solid-state analogue of the eutectic: a single solid phase decomposes
on cooling into two new solid phases, $\gamma \to \alpha + \beta$. The
iron-carbon eutectoid at $727\,\degC$ and $0.76\,\%$ carbon, where
austenite decomposes to ferrite plus cementite, is the most
important eutectoid in materials engineering (Section~7 of this
chapter). The peritectoid, $\alpha + \beta \to \gamma$, is the
solid-state analogue of the peritectic.

\engterm{Intermediate compounds}. Many binary systems contain
intermediate phases (sometimes called \engterm{intermetallic
compounds}) that exist over a narrow composition range and have
distinct crystal structures different from either pure component.
Cementite (Fe$_3$C) in iron-carbon, brass intermetallics in
copper-zinc ($\beta$, $\gamma$, $\delta$, $\varepsilon$ phases), and
the silver-copper system's $\beta$ phase are typical. Intermediate
phases are usually brittle (because their crystal structures often
lack the slip systems of pure metallic lattices) and contribute
hardness to alloys whose microstructure contains them.

\section{Ternary diagrams}\pathtag{mastery}

A ternary phase diagram displays three-component systems on a triangular
composition plane (Gibbs triangle), with temperature added as a third
dimension. Three-component systems are pervasive in real engineering:
all real steels are at least iron-carbon-manganese ternaries, real
aluminium alloys are at least aluminium-copper-magnesium ternaries,
and real solders are at least tin-silver-copper ternaries.

\engterm{Reading the Gibbs triangle}. The three pure components sit at
the vertices of an equilateral triangle. Any point inside the triangle
represents a composition with all three components present. The
distance from a given side is proportional to the fraction of the
opposite vertex's component. A point on a side represents a binary
composition with the third component absent. The three lines from a
single vertex to the opposite side along constant ratios of the other
two components are the \engterm{isopleths} for that vertex.

\engterm{Isothermal sections}. The most common engineering reading of
a ternary system is the \engterm{isothermal section}, the
intersection of the three-dimensional ternary diagram with a horizontal
($T = \text{const}$) plane. The section shows regions of single, two,
and three phases at the chosen temperature. Tie lines in two-phase
regions connect equilibrium compositions of the two phases; tie
triangles in three-phase regions have vertices at the three
equilibrium phase compositions, with the lever rule extended to two
dimensions to determine phase fractions.

\engterm{Liquidus surfaces}. The three-dimensional ternary diagram's
liquidus is a surface in $T$-$x_A$-$x_B$ space; projecting it onto the
composition triangle gives the \engterm{liquidus projection}, with
contours of constant liquidus temperature and lines showing primary
crystallisation paths. The projection is the working tool for casting
and solidification engineering: it identifies which solid phase
nucleates first as the liquid cools through the liquidus, and how the
liquid composition evolves as solidification proceeds along the
two-phase boundary.

The detailed reading of ternary diagrams is beyond the working depth
of this chapter; the engineering reader who needs to specify a casting
alloy or a brazing filler should consult the system-specific diagrams
in \textcite{text:porter-easterling-phase} or the ASM handbooks.

\section{Lever rule}\pathtag{core}

In a two-phase region of a binary diagram, the equilibrium compositions
of the two phases are read off the \engterm{tie line}: a horizontal
line at the alloy's temperature, intersecting the phase boundaries on
either side of the two-phase region. The fractions of the two phases
in the equilibrium mixture are determined by the \engterm{lever rule}:
the alloy's overall composition divides the tie line, and the fraction
of each phase equals the length of the opposite segment divided by the
total tie-line length.

\engterm{The rule as a balance}. Let $x_0$ be the alloy's overall
composition, $x_\alpha$ the composition of phase $\alpha$ at the tie
line, and $x_\beta$ the composition of phase $\beta$ at the tie line.
Mass conservation of the chosen component gives
\[
x_0 = f_\alpha x_\alpha + f_\beta x_\beta, \quad f_\alpha + f_\beta = 1,
\]
which solves to
\[
f_\alpha = \frac{x_\beta - x_0}{x_\beta - x_\alpha}, \quad f_\beta =
\frac{x_0 - x_\alpha}{x_\beta - x_\alpha}.
\]
The names ``lever rule'' and ``inverse lever rule'' both appear in the
literature; the form given is the most useful for engineering
calculation.

\engterm{Use cases}. The lever rule is the working tool for:

\begin{itemize}[leftmargin=*]
\item Predicting the fraction of liquid remaining as a cast alloy
  cools through the two-phase $L + \alpha$ region. The casting's
  shrinkage and porosity depend on this fraction.
\item Predicting the fraction of austenite remaining at a given
  temperature in the iron-carbon system during heat treatment, which
  governs the available carbon for the subsequent decomposition.
\item Predicting the fraction of liquid soldering alloy at a chosen
  reflow temperature, which governs the wetting behaviour of the
  joint.
\item Predicting the fraction of precipitate phase in an age-hardened
  aluminium alloy from the alloy's equilibrium composition at the
  ageing temperature.
\end{itemize}

\engterm{The non-equilibrium correction}. The lever rule assumes
infinitely slow cooling, with full diffusional homogenisation at each
temperature. Real casting and heat-treatment processes cool too
quickly for diffusional equilibrium in the solid; the actual phase
fractions and compositions follow the Scheil-Gulliver equation, which
accounts for the lack of diffusion in the solid while still allowing
diffusion in the liquid. The Scheil equation predicts more eutectic
phase fraction than the lever rule and a wider freezing range, both of
which are observed in practice.

\begin{principle}[The lever rule is a balance]
The lever rule for tie-line compositions in a two-phase region is
mass conservation applied to a single component across the two
equilibrium phases. The discipline is the same one used for force
balances in statics, energy balances in thermodynamics, and species
balances in reaction engineering. The reader who can write the
balance and solve for the two phase fractions reads off any phase
diagram in the materials-engineering literature.
\end{principle}

\section{Eutectics, peritectics, and microstructural consequences}\pathtag{standard}

The invariant reactions of binary systems (eutectic, peritectic,
eutectoid, peritectoid) produce characteristic microstructures whose
features are read off the diagram. The eutectic microstructure of a
slowly cooled lead-tin eutectic is a fine lamellar interleaving of
$\alpha$ (Pb-rich) and $\beta$ (Sn-rich) phases. A
\engterm{hypoeutectic} composition (less tin than the eutectic
composition) shows primary $\alpha$ dendrites embedded in a eutectic
matrix; a \engterm{hypereutectic} composition (more tin) shows primary
$\beta$ in a eutectic matrix. The fraction of eutectic vs primary
phase is given by the lever rule applied at the eutectic temperature.

\engterm{Microstructural scale and properties}. The spacing of the
eutectic lamellae depends inversely on the cooling rate: faster
cooling produces finer lamellae and harder microstructure (because the
finer scale of two-phase interfaces resists dislocation motion). The
Hall-Petch-like relation between hardness and lamellar spacing is
empirical and material-specific. For lead-tin eutectic, the lamellar
spacing in slow-cooled service castings is $\sim 1$ to $5\,
\si{\micro\meter}$; in fast-cooled wave-solder joints, $\sim 0.1$ to
$0.3\,\si{\micro\meter}$. The harder, finer microstructure has
higher fatigue resistance, which matters for the thermal-cycling
service of solder joints in electronic equipment.

\engterm{Peritectic microstructures}. Peritectic alloys show
characteristic core-shell structures: a core of the original solid
phase, a shell of the peritectic product, and unreacted liquid that
solidifies in the inter-grain space. Long homogenisation anneals
diffuse out the cores in slow-grown crystals; in service castings the
cores often persist and produce inhomogeneous behaviour.

\engterm{Solid-state precipitation}. A solid phase that becomes
supersaturated on cooling can precipitate a second phase, with the
new phase nucleating at the grain boundaries (heterogeneous
nucleation, favoured at high temperatures with slow cooling) or
throughout the grain volume (homogeneous nucleation, favoured at low
temperatures with rapid cooling and large supersaturation). The
microstructure of the precipitate (its size, distribution, and
crystallographic orientation relative to the matrix) sets the
strengthening contribution. Chapter~5 of this volume develops
precipitation strengthening as one of the principal mechanisms for
high-strength alloys.

\section{Phase transformations and TTT diagrams}\pathtag{standard}

A phase diagram tells what is stable; a \engterm{time-temperature-
transformation diagram} (TTT) tells how fast the system gets there.
The TTT diagram of an alloy at a chosen starting composition shows
the time required, at each temperature, to begin and to complete the
diffusional transformation from the starting phase to the equilibrium
phase. The diagram has the characteristic ``C'' shape: at high
temperatures the driving force for transformation is small and
nucleation is slow; at low temperatures the driving force is large
but diffusion is slow; at intermediate temperatures (the ``nose'' of
the C) both rates are moderate and transformation is fastest.

\engterm{TTT for iron-carbon}. The iron-carbon TTT diagram is the
foundation of every steel heat treatment. Starting from an austenite
($\gamma$-iron with dissolved carbon) at $850\,\degC$, the alloy can
transform to:

\begin{itemize}[leftmargin=*]
\item \engterm{Pearlite}: lamellar mixture of ferrite and cementite,
  formed by slow cooling (or isothermal hold at $600$ to $700\,\degC$).
\item \engterm{Bainite}: fine non-lamellar mixture of ferrite and
  cementite, formed by intermediate cooling (or isothermal hold at
  $250$ to $550\,\degC$).
\item \engterm{Martensite}: supersaturated body-centred-tetragonal
  iron with carbon trapped in solid solution, formed by rapid
  quenching that bypasses both pearlite and bainite formation. The
  martensite-start temperature $M_s$ depends on the alloy's carbon
  content and other alloying elements; for plain-carbon $0.8\,\%$ C
  steel, $M_s \approx 220\,\degC$.
\end{itemize}

The selection among the three transformation products is by cooling
path. The Davenport-Bain isothermal-transformation experiments
\cite{hist:bain1939} from the US Steel laboratories in the late
1920s established the TTT diagram as the working tool of the
metallurgist. The continuous-cooling-transformation (CCT) diagram is
a related construction that uses continuous cooling paths rather than
isothermal holds; CCT diagrams are more directly useful for industrial
heat-treatment furnaces, in which the workpiece typically cools
continuously rather than soaking at fixed temperature.

\engterm{Hardenability}. The depth to which a quenched steel section
forms martensite is set by the steel's hardenability, which depends
on the cooling rate at depth (governed by the quench medium and the
section's thermal mass) and on the alloying elements that shift the
TTT diagram's nose to longer times. Manganese, molybdenum, chromium,
and nickel all slow the pearlite-and-bainite transformations and
thereby increase hardenability. The Jominy end-quench test is the
standard industrial measurement: a $25\,\text{mm}$ diameter bar is
heated to austenitising temperature, end-quenched with a controlled
water spray, and hardness measured along its length. The plot of
hardness versus distance from the quenched end is the steel's
hardenability profile.

\engterm{Martensitic transformation}. The martensite-formation
reaction is diffusionless: the FCC austenite lattice transforms to a
body-centred-tetragonal arrangement by a coordinated shear motion
(the Bain strain) without atomic diffusion. The reaction proceeds at
the speed of sound in the lattice; the transformation does not
nucleate at a temperature, it nucleates at a strain. The
\engterm{martensite-start temperature} $M_s$ is the temperature below
which the driving force exceeds the strain-energy barrier. Holding
isothermally below $M_s$ does not produce more martensite; cooling to
a lower temperature does. The martensite is hard ($\sim 700$ to $900\,
\text{HV}$ for high-carbon martensite) and brittle (the trapped
carbon produces lattice distortion that resists dislocation motion).
Tempering at intermediate temperatures ($200$ to $600\,\degC$) allows
some carbon to precipitate as cementite, relieving the lattice
distortion and trading hardness for ductility. The
quench-and-temper cycle is the working heat treatment for nearly
every structural steel.

\section{Worked example: Fe-C and steel heat treatment}\pathtag{core}

The iron-carbon phase diagram is the most studied and most
engineering-significant phase diagram. Iron and carbon do not form a
true equilibrium two-phase system; the carbon is metastable as
cementite (Fe$_3$C) but slowly converts to graphite in service at
high temperatures. The metastable Fe-Fe$_3$C diagram is the working
diagram for steel; the stable Fe-graphite diagram is used for cast
irons.

\engterm{The key features}. The diagram on the composition axis runs
from pure iron to $6.67\,\%$ carbon (pure Fe$_3$C). The key features:

\begin{itemize}[leftmargin=*]
\item \engterm{$\alpha$-ferrite}: BCC iron with very limited carbon
  solubility (maximum $0.022\,\%$ at $727\,\degC$). Soft, ductile,
  magnetic below $770\,\degC$.
\item \engterm{$\gamma$-austenite}: FCC iron with carbon solubility up
  to $2.14\,\%$ at $1147\,\degC$. The phase from which heat
  treatments begin.
\item \engterm{Cementite Fe$_3$C}: orthorhombic intermetallic with
  $6.67\,\%$ carbon by mass. Hard ($\sim 800\,\text{HV}$), brittle,
  metastable with respect to ferrite plus graphite at long times.
\item \engterm{Pearlite}: the eutectoid mixture of $\alpha$-ferrite
  and cementite formed by the reaction $\gamma \to \alpha +
  \text{Fe}_3\text{C}$ at $727\,\degC$ and $0.76\,\%$ carbon. Lamellar
  microstructure, hardness $\sim 200\,\text{HV}$.
\item The eutectic point at $4.30\,\%$ carbon and $1147\,\degC$:
  liquid solidifies to austenite plus cementite. This is the
  high-carbon range of the cast irons.
\end{itemize}

\engterm{Steel classifications}. Plain-carbon steels with less than
$0.76\,\%$ carbon are \engterm{hypoeutectoid}, between $0.76\,\%$ and
$2.14\,\%$ are \engterm{hypereutectoid}, between $2.14\,\%$ and $6.67\,
\%$ are \engterm{cast irons} (sub-eutectic, eutectic, and
hyper-eutectic). The American Iron and Steel Institute (AISI)
designations and SAE numbers identify common compositions: AISI 1018
is $0.18\,\%$ carbon plain-steel, AISI 1080 is $0.80\,\%$ near-
eutectoid, AISI 4340 is a low-alloy steel with $0.4\,\%$ C, Ni, Cr,
and Mo.

\engterm{Annealing}. Slow furnace cooling of an austenitised steel
produces coarse pearlite at the eutectoid composition, or
proeutectoid ferrite plus pearlite for hypoeutectoid steels, or
proeutectoid cementite plus pearlite for hypereutectoid steels. The
microstructure is soft and ductile; annealing is used to relieve
internal stresses and to prepare material for cold working.

\engterm{Normalising}. Air cooling from austenitising temperature
produces finer pearlite and finer grain size than annealing. The
microstructure is harder and stronger than annealed material;
normalising is the standard finish for many structural-steel
components.

\engterm{Quenching}. Rapid cooling (typically in water, oil, or
polymer-water quench solutions) suppresses both pearlite and bainite
and produces martensite. As-quenched martensite is hard and brittle;
it is essentially never used as-quenched in structural service.

\engterm{Tempering}. Reheating quenched martensite to a temperature
between $200$ and $600\,\degC$ for tens of minutes to hours allows
some carbon to precipitate as cementite. The hardness decreases as
the tempering temperature increases; the toughness increases. The
choice of tempering temperature is the engineer's trade-off between
strength and toughness for the application. High-strength fasteners
are tempered to $\sim 425\,\degC$; tool steels to higher temperatures
that produce secondary hardening from alloy carbides.

\engterm{The quench-and-temper cycle in practice}. The reader who
heat-treats a small piece of $0.4\,\%$ C steel:

\begin{enumerate}[leftmargin=*]
\item Austenitise: heat to $850\,\degC$ and hold for $30\,\text{min}$
  per $25\,\text{mm}$ of section thickness, to dissolve all carbon
  into the FCC austenite.
\item Quench: transfer rapidly to a water or oil bath at room
  temperature. The cooling rate must exceed the critical rate (the
  rate at which the CCT diagram's nose is just missed) to produce
  fully martensitic structure. Water quenching produces fast cooling
  with risk of distortion or cracking; oil quenching is slower and
  gentler.
\item Temper: heat to the chosen tempering temperature (e.g. $425\,
  \degC$) for $30\,\text{min}$ to $2\,\text{h}$, then air cool. The
  microstructure becomes tempered martensite, a fine dispersion of
  cementite particles in a ferrite matrix, with hardness $\sim 45$ to
  $50\,\text{HRC}$ for $0.4\,\%$ C steel.
\end{enumerate}

The chapter's project repeats the cycle in practice and measures the
resulting hardness.

\section{Failure: improper heat treatment and brittle parts}\pathtag{core}

Improper heat treatment of structural steels produces components
whose service performance differs sharply from the design intent.
The failure mode is widespread because heat treatment is a process
step whose outcome is not visible to ordinary inspection: a properly
tempered $4140$ shaft and a partially-tempered or untempered
$4140$ shaft look identical. The differences emerge only under load.

\engterm{Incomplete austenitisation}. If the workpiece is undersoaked
(insufficient time at austenitising temperature, or temperature too
low), the carbide phase does not fully dissolve into the austenite.
On quenching, the dissolved carbon produces martensite while the
undissolved carbides remain. The resulting microstructure is a
mixture of martensite and pearlite or ferrite, with hardness lower
than expected and inhomogeneous from point to point. The component
is softer than design specification in some locations and harder in
others; under cyclic load, the soft regions deform preferentially
and concentrate strain at the boundary with the harder regions.

\engterm{Quench cracking}. Rapid quenching of a steel section in
water produces high thermal stresses and high transformation
stresses (martensite has a larger volume than austenite by $\sim 4\,
\%$). The combination can crack the workpiece, particularly at
sharp corners, holes, and section transitions where the stress
concentration is highest. Quench cracks are typically subsurface
and may not be detected by surface inspection; magnetic-particle or
ultrasonic inspection is required. Quench cracking is the classic
failure mode of inexpensive water-quenched parts; oil quenching and
polymer-quench solutions reduce the risk at the cost of slower
cooling.

\engterm{Untempered or undertempered martensite}. Untempered
martensite is extremely brittle. A workpiece quenched to martensite
and not subsequently tempered has hardness as high as $700\,\text{HV}$
($\sim 60\,\text{HRC}$) but a tensile fracture strain of less than
$1\,\%$. Under any significant tensile or bending load, the part
fails by brittle fracture along prior austenite grain boundaries.
The failure surface is characteristically intergranular and shows
no plastic deformation. Untempered martensite is the cause of
through-cracking failures in shafts, gears, and fasteners that
``should not'' have failed at their measured static load. The
mitigation is straightforward: temper after quench, always, and
verify the hardness against the specification.

\engterm{Tempered martensite embrittlement (TME)}. Tempering in the
$250$ to $400\,\degC$ range produces a specific form of embrittlement
in some alloy steels (notably those containing phosphorus,
antimony, or tin impurities). The embrittling mechanism is the
segregation of the impurity element to prior austenite grain
boundaries, where it weakens the bond between adjacent grains. The
toughness drops sharply in this temperature range and recovers at
higher temperatures. The engineering response is to specify
tempering temperatures outside the embrittlement range and to limit
the residual concentrations of the embrittling impurities at the
steelmaking stage.

\engterm{Hydrogen embrittlement during heat treatment}. Steels
exposed to hydrogen during austenitising or quenching (from moist
atmospheres, from oil-based quench media, from plating operations
that follow heat treatment) absorb hydrogen at the atomic level. The
absorbed hydrogen segregates to dislocation cores and grain
boundaries, where it embrittles the structure. High-strength steels
($> 1000\,\text{MPa}$ tensile strength) are especially susceptible.
The standard mitigation is a low-temperature baking treatment
(typically $190\,\degC$ for $24\,\text{h}$) immediately after
plating or after high-temperature exposure in a hydrogen-bearing
atmosphere, to drive the hydrogen out of the lattice before service.
The mitigation must be done within hours of the hydrogen exposure;
once the hydrogen has reached critical concentrations at
crack-initiation sites, the part is committed to brittle failure
under load. Chapter~5 of this volume develops hydrogen embrittlement
further.

\engterm{Failure pattern}. A repeated pattern in heat-treatment-related
component failures: the component is correctly designed for static
load, the heat treatment is partially or improperly executed (often
in a job-shop context with insufficient process control), and the
component fails in service under a load well within its nominal
capacity. The brittle nature of the failure and the absence of
plastic warning means the operator typically gets no notice. The
engineering response includes specification of heat-treatment process
controls (atmosphere, time, temperature, quench medium, post-
tempering hardness verification), audit of the heat-treatment
supplier's records, and inspection of finished components for hardness
and microstructure on a sampling basis. The standard certification
documents (mill test reports, heat-treatment certificates) are the
engineering paper trail that protects against the failure mode.

\engterm{Related failure: thermal-history-induced material degradation
in service}. Components that experience temperatures outside their
heat-treatment design envelope during service undergo unintended
phase transformations. Boiler tubes that overheat above their design
temperature transform from the as-fabricated tempered-martensite
structure to a softer ferrite plus coarse carbide; the tubes lose
creep strength and rupture under operating pressure. The
Three Mile Island unit~2 accident of 1979 \cite{acc:kemeny-tmi-1979}
included a sequence in which a partial core uncovering led to fuel
cladding (Zircaloy-4) reaching temperatures at which the
$\alpha$-to-$\beta$ phase transition of zirconium occurred and the
steam-zirconium oxidation reaction became rapid. The TMI mechanism is
not primarily heat-treatment-induced (the cladding was correctly
manufactured); it is phase-diagram-related in that the cladding's
service environment exceeded the phase-stability envelope of the
material as designed. The wider engineering pattern is the same:
material whose service temperature exceeds the design envelope
undergoes phase changes that degrade its properties, and the design
envelope must include the worst-case fault scenarios as well as
nominal operation.

\begin{principle}[Phase diagrams gate heat treatment]
The heat treatment of any alloy follows a path on the phase diagram.
The properties of the finished component reflect the path: the
phases present at each temperature, the fraction of each phase, the
microstructural scale and morphology set by cooling rate. A heat
treatment that deviates from the design path (under-austenitisation,
slow quench, missed tempering step, wrong tempering temperature)
produces a different microstructure and different properties. The
engineering content of heat-treatment specification is the
documentation of the path, the verification of each step, and the
hardness-or-microstructure check on the finished component.
\end{principle}

\begin{estimation}
\textbf{Question.} Estimate the hardness of a $0.45\,\%$ C plain-
carbon steel after water quenching from $850\,\degC$, and again
after tempering at $400\,\degC$ for $1\,\text{h}$.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The water-quenched microstructure is martensite. The hardness of as-
quenched martensite scales with carbon content; for plain-carbon
steel the empirical relation is approximately
\[
H_{\text{martensite}} \approx 500 + 800\,(\%\text{C})^{1/2}
\quad \text{(HV)}.
\]
For $0.45\,\%$ C: $H \approx 500 + 800 \times 0.671 \approx 1040\,
\text{HV}$. Converting to Rockwell C, $1040\,\text{HV} \approx 65\,
\text{HRC}$.

After tempering at $400\,\degC$ for $1\,\text{h}$, the hardness drops
because some martensite decomposes to fine cementite plus ferrite.
The tempering relation for plain-carbon steel is approximately a
$50\,$-$\text{HV}$ drop per $100\,\degC$ tempering temperature above
room temperature; at $400\,\degC$ tempering this gives a drop of
$\sim 200\,\text{HV}$, leaving $\sim 840\,\text{HV} \approx 55\,
\text{HRC}$. Experimental data for $0.45\,\%$ C steel give $\sim 470\,
\text{HV}$ ($\sim 48\,\text{HRC}$) at this tempering temperature, so
our rule of thumb overestimates the residual hardness by $\sim 35\,\%$.

\smallskip
\textit{Postmortem.} The estimate of as-quenched hardness is within
$10\,\%$ of measured values for plain-carbon steels in the $0.3$ to
$0.6\,\%$ C range. The tempering estimate is poor: the actual
softening at $400\,\degC$ is faster than the linear rule suggests
because the precipitation of $\varepsilon$-carbide and the
transformation of retained austenite both happen in this temperature
range. A better estimate uses the Hollomon-Jaffe tempering parameter
$P = T(20 + \log t)$ (with $T$ in kelvin and $t$ in hours) and
empirical tempering curves for the specific alloy. The chapter's
project uses small-scale measurement to extract the actual hardness;
the estimate above sets the order of magnitude.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Heat-treat a steel sample and measure the hardness]
The reader heat-treats three small samples of $0.4$ to $0.6\,\%$ C
plain-carbon steel through three different cooling paths and measures
the hardness of each. The samples are interpreted on the iron-carbon
phase diagram.

\textbf{Track}. Physical heat treatment.
\textbf{Hazard class}: moderate. Small-furnace temperatures up to
$900\,\degC$; quenching of hot metal in water or oil; safety glasses,
heat-resistant gloves, and tongs required. The work is appropriate
for a school laboratory or a maker-space metalworking station.

\textbf{Materials}.

\begin{itemize}[leftmargin=*]
\item Three small samples of plain-carbon steel of approximately
  $0.45\,\%$ C (typical drill-rod or hot-rolled bar stock, $\sim 20\,
  \text{mm}$ diameter, $\sim 30\,\text{mm}$ length). The samples must
  be of identical material; the variable is the heat-treatment path.
\item A small electric muffle furnace capable of reaching $900\,
  \degC$.
\item A quench bath of water at room temperature; a separate quench
  bath of oil (vegetable or mineral) at room temperature.
\item A hardness tester (Rockwell C, Brinell, or Vickers; a digital
  portable hardness tester is acceptable if calibrated).
\end{itemize}

\textbf{Procedure}.

\begin{enumerate}[leftmargin=*]
\item Mark each sample with an identifier. Measure the as-received
  hardness of one sample. This is the baseline ``hot-rolled'' value.
\item Heat all three samples in the furnace to $850\,\degC$ and hold
  for $30\,\text{min}$.
\item Cool one sample slowly in the furnace (turn furnace off,
  leaving the sample inside until temperature falls below $300\,
  \degC$). Label this sample ``annealed.''
\item Remove the second sample from the furnace and air-cool on a
  refractory plate. Label this sample ``normalised.''
\item Remove the third sample and immediately immerse it in the
  water-quench bath, agitating to break the vapour blanket. Hold
  underwater until cool. Label this sample ``quenched.''
\item Measure the hardness of each of the three samples at three
  locations on the surface (after cleaning off scale with abrasive).
  Record the mean and the range.
\item Cut the quenched sample longitudinally and reheat one half to
  $400\,\degC$ for $1\,\text{h}$, then air cool. This half is
  ``quenched and tempered.''
\item Measure the hardness of the quenched-and-tempered half and
  compare to the quenched-only half.
\end{enumerate}

\textbf{Analysis}.

\begin{itemize}[leftmargin=*]
\item Plot hardness versus cooling rate for the four conditions.
\item Identify the cooling rate of each cooling path qualitatively
  (slow, intermediate, fast) and locate the corresponding
  microstructure on the iron-carbon TTT diagram.
\item Estimate the tempering parameter for the quenched-and-tempered
  sample and predict the hardness from a standard tempering chart.
\item Compare measured to predicted hardness for the four conditions.
\end{itemize}

\textbf{Deliverable}. A six- to ten-page report including the data
table, the plot, the TTT-diagram annotation, and a paragraph on the
discrepancies between predicted and measured hardness.

\textbf{Time}. Four to six hours of bench work (including furnace
heat-up and cool-down), two to three hours of write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
The triple point of water is at $T_3 = 0.01\,\degC$ and $p_3 =
611.66\,\text{Pa}$. Use the Clausius-Clapeyron equation with
$\Delta H_{\text{vap}} = 45\,\text{kJ}/\text{mol}$ to compute the
vapour pressure of water at $25\,\degC$ and at $80\,\degC$. Compare
to the tabulated values.
\end{exercise}

\begin{exercise}
For a copper-nickel alloy of overall composition $40\,\%$ Ni at
$1300\,\degC$, use the Cu-Ni phase diagram (liquidus at $40\,\%$
Ni $\approx 1270\,\degC$, solidus at $40\,\%$ Ni $\approx 1220\,\degC$)
to apply the lever rule and compute the mass fractions of solid and
liquid. Compute the composition of each phase.
\end{exercise}

\begin{exercise}
A lead-tin alloy of composition $40\,\%$ Sn is cooled from the liquid
through the two-phase $L + \alpha$ region. Compute the fraction of
liquid remaining just above the eutectic temperature ($183\,\degC$).
Use the phase boundaries $L \approx 60\,\%$ Sn and $\alpha \approx
19\,\%$ Sn at $183\,\degC$.
\end{exercise}

\begin{exercise}
A $0.45\,\%$ C plain-carbon steel is slowly cooled from $850\,\degC$
to room temperature. Use the iron-carbon phase diagram to compute the
mass fraction of proeutectoid ferrite, pearlite, and cementite in the
final microstructure.
\end{exercise}

\begin{exercise}
A binary system has eutectic at $400\,\degC$ and $50\,\%$ B. The
$\alpha$ solidus at the eutectic temperature is at $5\,\%$ B; the
$\beta$ solidus is at $90\,\%$ B. Compute the fraction of eutectic
microconstituent in an alloy of composition $30\,\%$ B at $399\,
\degC$.
\end{exercise}

\begin{exercise}
The Hollomon-Jaffe tempering parameter is $P = T(20 + \log t)$ with
$T$ in kelvin and $t$ in hours. Compute the parameter for tempering
at $200\,\degC$ for $2\,\text{h}$ and for $400\,\degC$ for $0.5\,\text{h}$.
Identify which tempering condition produces softer steel.
\end{exercise}

\begin{exercise}
A steel with $M_s = 200\,\degC$ and $M_f = 50\,\degC$ is quenched to
$100\,\degC$ from austenitising temperature. Estimate the fraction of
martensite formed using the empirical Koistinen-Marburger relation
$f_M = 1 - \exp[-0.011(M_s - T)]$ for $T < M_s$.
\end{exercise}

\begin{exercise}
For a ternary aluminium-copper-magnesium alloy of composition
$\text{Al}\text{-}4\%\text{Cu}\text{-}1.5\%\text{Mg}$, use a Gibbs
triangle to identify the alloy's location and identify which primary
phase nucleates first on cooling from the liquid. Consult Reference
\cite{text:porter-easterling-phase} or a published Al-Cu-Mg liquidus
projection.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the Clausius-Clapeyron equation $dp/dT = \Delta H_{\text{tr}}
/(T \Delta V_{\text{tr}})$ from the equilibrium condition that the
Gibbs free energies of the two phases are equal along the phase
boundary.
\end{exercise}

\begin{exercise}
Derive the lever rule $f_\alpha = (x_\beta - x_0)/(x_\beta - x_\alpha)$
from mass conservation of one component across the two phases.
\end{exercise}

\begin{exercise}
Derive the Gibbs phase rule $F = C - P + 2$ from a count of intensive
variables and equilibrium constraints for a system of $C$ components
in $P$ phases.
\end{exercise}

\begin{exercise}
Show that the eutectic-microstructural mass fraction in a hypoeutectic
binary alloy equals $(x_0 - x_\alpha^{\text{eut}})/(x_{\text{eut}} -
x_\alpha^{\text{eut}})$, where $x_\alpha^{\text{eut}}$ is the
$\alpha$-phase composition at the eutectic temperature.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the boiling point of water at the summit of Mount Everest
($8848\,\text{m}$, ambient pressure $\sim 0.33\,\text{atm}$) using the
Clausius-Clapeyron equation.
\end{exercise}

\begin{exercise}
Estimate the order of magnitude of the cooling rate needed to produce
fully martensitic structure in a $25\,\text{mm}$ diameter rod of
$1\,\%$ C steel from austenitising temperature ($850\,\degC$).
\end{exercise}

\begin{exercise}
Estimate the time required for diffusional homogenisation of a $1\,
\si{\milli\meter}$-thick layer in iron at $1000\,\degC$, given the
self-diffusion coefficient of iron $D \sim 10^{-13}\,\text{m}^2/
\si{\second}$ at $1000\,\degC$. Use the scaling $t \sim L^2/D$.
\end{exercise}

\begin{exercise}
Estimate the volume change associated with the austenite-to-
martensite transformation in plain-carbon steel, given that the FCC
austenite lattice constant is $\sim 3.6\,\si{\angstrom}$ and the
BCT martensite is $\sim 2.85\,\si{\angstrom}$ with $c/a \approx 1.05$.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a Scheil-Gulliver solidification model for a binary
isomorphous system. Compare the predicted solid-fraction-versus-
temperature curve to the lever-rule prediction. Identify the
conditions under which the two predictions differ significantly.
\end{exercise}

\begin{exercise}
Implement a one-dimensional heat-conduction model for a steel rod
quenched in water. Compute the temperature-time profile at the
centre and at the surface. Overlay the cooling paths on a TTT
diagram for $0.45\,\%$ C steel and predict the final microstructure
at each location.
\end{exercise}

\begin{exercise}
Implement a Jominy end-quench simulation: a cylindrical bar
$25\,\text{mm}$ in diameter, $100\,\text{mm}$ in length, end-quenched
with water spray on one face. Predict the hardness profile from the
local cooling rate and the steel's CCT diagram. Compare to published
Jominy curves.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify the steel grade and heat-treatment cycle for a shaft that
must support $1.0\,\text{GPa}$ tensile stress at room temperature
without fracture and tolerate cyclic loading. Identify the alloy
elements that provide hardenability for the required section
thickness ($50\,\text{mm}$ diameter).
\end{exercise}

\begin{exercise}
Specify a solder alloy for an electronic-assembly reflow process
that must wet copper pads, melt below $230\,\degC$, and tolerate
thermal cycling between $-40\,\degC$ and $+125\,\degC$ for $1000$
cycles. Identify the candidate eutectic compositions and the
trade-offs.
\end{exercise}

\begin{exercise}
Specify the heat-treatment cycle for an aluminium-copper-magnesium
alloy ($2024$ alloy) component to maximise yield strength.
Identify the solution-treatment temperature, the quench medium, and
the artificial-ageing temperature and time.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A metallographic section of a quenched steel shows mixed regions of
martensite and pearlite. Identify the most likely cause and propose
two heat-treatment modifications that would produce fully
martensitic structure.
\end{exercise}

\begin{exercise}
A solder joint shows a microstructure with large primary tin
dendrites in a fine eutectic matrix. Identify the alloy composition
range (hypo- vs hyper-eutectic) and estimate the cooling rate from
the dendrite arm spacing.
\end{exercise}

\begin{exercise}
A casting of an aluminium-silicon alloy ($7\,\%$ Si) shows coarse
needles of primary silicon. Identify the modification that would
refine the silicon morphology and explain its effect in terms of the
phase diagram and nucleation thermodynamics.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A heat-treated steel bolt fails after one year of service under
nominal static load. The fracture is intergranular and shows no
plastic deformation. Identify the most likely metallurgical cause
and the heat-treatment audit step that would have detected it.
\end{exercise}

\begin{exercise}
A boiler tube ruptures under nominal operating pressure after
fifteen years of service. The metallography shows coarse spheroidal
carbides in a ferrite matrix instead of the as-fabricated tempered
martensite. Identify the service-side mechanism and propose an
inspection that would have predicted the failure.
\end{exercise}

\begin{exercise}
A turbine blade in a gas-turbine engine fails at a location of
elevated operating temperature. The microstructure shows
$\gamma'$-precipitates that have coarsened far beyond the
design-specified size. Identify the underlying material change and
propose a service-life-limit re-derivation.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A colleague argues that the iron-carbon phase diagram is an
``equilibrium'' diagram and therefore irrelevant to heat-treated
steels, which are always in non-equilibrium states. Write a one-
paragraph reply identifying the correct relationship between the
equilibrium diagram, the TTT diagram, and engineering heat treatments.
\end{exercise}

\begin{exercise}
A vendor proposes to replace a quenched-and-tempered $4340$ shaft
with a normalised plain-carbon $1045$ shaft of the same diameter to
save cost. Identify the property changes the substitution would
produce and the application-specific factors that would make the
substitution acceptable or unacceptable.
\end{exercise}

\begin{exercise}
A student argues that the eutectic point on a binary diagram is
``special'' because it is the lowest melting point of the alloy
system. Identify the qualification this claim requires (the eutectic
is special as the invariant three-phase point, not as the overall
minimum) and state when the eutectic is in fact the system's lowest
liquidus.
\end{exercise}
