\chapter{Bearings, lubrication, tribology}
\label{vol08:ch02}

\epigraph{Friction is the price we pay for being able to put things down.}{anonymous tribologist}

\chapmeta{Half-life: medium-life. Archetypes: transport (lubricant film), failure. Exercise target: 30. Project track: physical.}

\noindent
A \engterm{bearing} supports a rotating or sliding load while permitting relative motion. The engineering of bearings rests on \engterm{tribology}: the science of interacting surfaces in relative motion, including friction, wear, and lubrication. Every rotating machine has bearings; bearing failure is a recurrent cause of machine failure across every industry that builds rotating equipment. This chapter teaches the working tribology and the working bearing-engineering discipline. The standard reference is the Stachowiak-Batchelor \emph{Engineering Tribology} \cite{text:stachowiak-batchelor}; for bearing-life calculations, the SKF and Timken catalogues remain the working sources.

\section{Friction, the engineering view}\pathtag{core}

The classical model of friction goes back to Amontons (1699) and Coulomb (1781): the friction force is proportional to the normal force, independent of the contact area, independent of the sliding speed. The friction coefficient $\mu$ depends on the materials in contact. Working values: $\mu \approx 0.6$ for steel on steel dry, $\approx 0.15$ with grease, $\approx 0.04$ with full hydrodynamic film, $\approx 0.001$ for rolling-element bearings.

The classical model is incomplete. Modern tribology recognises:

\textbf{Static vs.\ kinetic friction.} The friction coefficient at the onset of sliding is higher than during continuous sliding. The difference produces stick-slip motion at low speeds.

\textbf{Speed dependence.} The Stribeck curve shows friction varying with sliding speed in lubricated contacts: high at low speeds (boundary lubrication), decreasing through mixed and hydrodynamic regimes, rising again with viscous shear losses.

\textbf{Material specificity.} The friction coefficient is a property of the pair of materials, not of either one alone. The same steel surface has different $\mu$ against steel, brass, PTFE, ceramic.

\textbf{Surface roughness.} Real surfaces are not flat at micrometre scale; contact occurs at asperities. The friction force is the sum of adhesion, asperity deformation, and abrasion. Real area of contact is much smaller than apparent area.

The engineering implication: friction is not a single number. The design discipline is to specify the operating regime (lubrication, speed, load) and use the appropriate friction model for that regime.

\section{Plain and rolling bearings}\pathtag{core}

Two main bearing families dominate engineering practice.

\subsection{Plain bearings}

A \engterm{plain bearing} (also called a journal bearing or sleeve bearing) has a shaft rotating in a cylindrical sleeve, separated by a lubricant film. The film supports the load; direct metal-to-metal contact is avoided in steady operation.

\textbf{Hydrodynamic.} The shaft's rotation drags lubricant into a converging wedge; the resulting pressure supports the load. The film thickness is typically $1$-$30\,\si{\micro\meter}$. Used in automotive engine crankshafts, large turbine bearings, ship propeller shafts. Low maintenance once the right lubricant is in place; long life if the film is maintained.

\textbf{Hydrostatic.} An external pump pressurises the lubricant. The film exists even at zero speed; used in precision machine tools, large telescopes, and any application where the shaft must be supported at start-up.

\textbf{Boundary-lubricated plain bearings.} Materials like oil-impregnated bronze (Oilite), PTFE composites, or polymer journals. Used in low-speed, low-load applications where a full film is not maintained.

\subsection{Rolling-element bearings}

A \engterm{rolling-element bearing} has rolling elements (balls or rollers) between an inner race and outer race. The rolling contact has very low sliding friction; the bearing's friction coefficient is approximately $0.001$ to $0.002$.

\textbf{Deep-groove ball.} The standard general-purpose bearing. Radial and modest axial loads.

\textbf{Angular contact ball.} The contact angle allows higher axial loads. Used in pairs for thrust-supporting applications.

\textbf{Cylindrical roller.} Higher radial load capacity than ball bearings. No axial-load capacity unless flanged.

\textbf{Tapered roller.} Axial and radial loads. Used in vehicle wheel bearings, gearbox shafts.

\textbf{Spherical roller.} Self-aligning under shaft deflection or misalignment.

\textbf{Needle roller.} Compact; high radial load in a small package.

\textbf{Thrust bearings.} Axial loads only. Ball or roller variants.

The choice among rolling-element bearing types is workload-dependent: load magnitude, load direction (radial, axial, combined), speed, accuracy, life, cost.

\subsection{Comparison: plain vs.\ rolling}

\textbf{Plain bearings.} Higher load capacity per unit area, quieter, can run at very high speeds with proper lubrication, tolerant of contamination, generally cheaper for one-off large sizes. Higher friction at start-up; risk of seizure if lubrication fails.

\textbf{Rolling bearings.} Lower friction at start-up and at moderate speeds, standardised sizes, well-characterised life, easy to install. Limited by fatigue (a rolling-element bearing has a finite life under cyclic stress); sensitive to contamination and misalignment; noisier.

\section{Lubrication regimes: boundary, mixed, hydrodynamic, EHL}\pathtag{core}

The lubrication regime determines what separates the moving surfaces.

\subsection{The Stribeck curve}

The friction coefficient $\mu$ plotted against the dimensionless parameter $\eta U / P$ (viscosity times sliding speed divided by load per unit area) traces the Stribeck curve. The regimes:

\textbf{Boundary lubrication.} $\eta U / P$ very small. Surfaces in direct contact at asperities; lubricant chemistry (anti-wear additives, friction modifiers) carries the load between asperities. $\mu \approx 0.05$-$0.15$.

\textbf{Mixed lubrication.} Some asperity contact, some film. $\mu$ decreases as load is shared.

\textbf{Elastohydrodynamic lubrication (EHL).} The film is thin enough that the elastic deformation of the surfaces matters; the contact pressure modifies the lubricant viscosity (piezoviscous effect). Operates in rolling-element bearings and gear contacts. Film thickness on the order of $0.1$-$1\,\si{\micro\meter}$. $\mu \approx 0.005$-$0.02$.

\textbf{Hydrodynamic lubrication.} Full film separates the surfaces; no asperity contact. $\mu \approx 0.001$-$0.01$. Operates in well-designed journal bearings.

\textbf{Hydrostatic.} External pressure maintains the film independent of speed.

The engineering discipline is to design the bearing for the intended regime, then choose the lubricant and operating conditions that keep it there.

\subsection{Film thickness}

The Reynolds equation governs the lubricant film. For a hydrodynamic journal bearing, the central film thickness is approximately
\[
h_0 \approx \left( \frac{6 \eta U L}{P} \right)^{1/2}
\]
for some characteristic geometry. The engineering rule: the minimum film thickness must exceed the combined surface roughness of the two contacts by a factor of about 3 ($\lambda$ ratio $> 3$) for full-film hydrodynamic operation.

\subsection{The temperature problem}

Lubricant viscosity drops with temperature. A bearing's heat generation depends on speed and friction; the bearing's heat dissipation depends on the housing and the lubricant flow. The steady-state operating temperature is the solution of an energy balance. Working bearings run at $40$-$80\,\si{\degreeCelsius}$; the lubricant's viscosity at that temperature must support the film.

\section{Lubricant chemistry}\pathtag{standard}

A lubricant is a fluid (mostly) chosen to maintain the film, manage temperature, and protect the surfaces.

\subsection{Composition}

\textbf{Base oil.} Mineral oil, synthetic ester, polyalphaolefin, polyalkylene glycol. The base oil sets viscosity, temperature range, and chemical stability.

\textbf{Additives.} Anti-wear (zinc dialkyldithiophosphate, ZDDP), extreme-pressure (sulphur, phosphorus compounds), antioxidants, viscosity-index improvers, friction modifiers, detergent and dispersants for engine oils, anti-corrosion. Modern lubricants are 10-30\% additive by mass.

\textbf{Greases.} Lubricant trapped in a soap or polymer thickener. Hold the lubricant in place; used where re-lubrication is impractical.

\subsection{Viscosity grading}

The standard scales: SAE for engine and gear oils (0W-20, 10W-40, 75W-90, etc.); ISO VG for industrial oils (VG 32, VG 68, VG 220, etc.). The first number indicates low-temperature behaviour; the second high-temperature.

\subsection{Lubricant degradation}

Lubricants degrade in service: oxidation, water contamination, particulate contamination, additive depletion, viscosity change. The engineering response: periodic oil analysis (viscosity, total acid number, particle count, wear-metal content), filtration, scheduled change intervals. Modern industrial practice uses oil-condition monitoring to extend change intervals beyond fixed schedules.

\section{Wear and life calculations}\pathtag{core}

\subsection{Bearing life (rolling-element)}

The standard \engterm{$L_{10}$ life} is the number of revolutions (or hours) at which 90\% of identical bearings have not yet failed by fatigue. The Lundberg-Palmgren equation:
\[
L_{10} = \left( \frac{C}{P} \right)^p
\]
where $C$ is the dynamic load capacity (from manufacturer's data), $P$ is the actual load, $p = 3$ for ball bearings, $p = 10/3$ for roller bearings. The life is in millions of revolutions; multiply by speed (in rpm) and divide by 60 to get hours.

\textbf{Modified life ($L_{na}$).} Modern bearing-life calculations modify $L_{10}$ for reliability requirement, lubrication regime, and contamination. The ISO 281 modified life:
\[
L_{na} = a_1 a_{\mathrm{ISO}} L_{10}
\]
where $a_1$ adjusts for reliability and $a_{\mathrm{ISO}}$ adjusts for lubrication. A well-lubricated bearing in clean conditions has $a_{\mathrm{ISO}} \gg 1$; a contaminated, poorly-lubricated bearing has $a_{\mathrm{ISO}} \ll 1$ and the bearing can fail orders of magnitude before $L_{10}$.

\subsection{Wear mechanisms}

\textbf{Adhesive wear.} Material transfer at asperity contacts. Increases with load and sliding distance; minimised by film lubrication.

\textbf{Abrasive wear.} Hard particles plough through softer surfaces. Mitigated by surface hardening and by filtration.

\textbf{Fatigue wear.} Sub-surface crack initiation under repeated contact; the dominant rolling-element-bearing failure mode.

\textbf{Corrosive wear.} Surface reaction with the environment, followed by removal of the reacted layer.

\textbf{Erosive wear.} Solid particles in a fluid stream impact and remove surface material.

\textbf{Fretting.} Small-amplitude oscillating motion between surfaces; produces wear faster than the displacement alone would predict.

The engineering response is to identify the dominant wear mechanism for the specific application and design accordingly: material selection, surface treatment, lubrication regime, sealing.

\section{Seals and gaskets}\pathtag{standard}

A \engterm{seal} prevents fluid (or particulate) from crossing a boundary. The standard types:

\textbf{Static seals.} O-rings, gaskets, sealants for non-moving joints. Sized for the bore and the pressure; material chosen for chemical compatibility with the fluid.

\textbf{Lip seals.} For rotating shafts. A elastomeric lip rides on the shaft; a garter spring maintains contact pressure. Limited speed and pressure capability; widely used for moderate-duty rotating shafts.

\textbf{Mechanical seals.} Two flat faces, one fixed and one rotating, pressed together. Used for high-pressure, high-speed shaft seals in pumps and compressors.

\textbf{Labyrinth seals.} Non-contact; rely on flow resistance through a tortuous path. Used in high-speed turbomachinery.

\textbf{Magnetic fluid seals.} Ferrofluid held in a magnetic field. Used in vacuum and clean-room applications.

The engineering choice is pressure, speed, fluid, leak tolerance.

\section{Failure: lubrication failures, premature bearing death}\pathtag{core}

A bearing that achieves its $L_{10}$ life has performed as designed. A bearing that fails substantially before $L_{10}$ has experienced a discrete failure mode. The canonical patterns:

\subsection{Lubrication failure}

\textbf{Lubricant depletion.} Grease dries out, oil leaks, lubricant volume drops below the bearing's needs. The film disappears; metal-to-metal contact begins; the bearing welds and seizes within minutes to hours.

\textbf{Contamination.} Hard particles (dust, swarf, wear debris) enter the bearing. Particles are forced through the rolling contact, indenting the raceway. Each indent becomes a stress raiser and a future fatigue initiation site. The bearing's effective life is reduced by orders of magnitude.

\textbf{Water contamination.} Water in lubricant reduces film thickness, accelerates corrosion, and (in lithium soap greases) breaks down the thickener. A bearing in a wet environment without proper sealing has substantially reduced life.

The engineering response: design the lubrication system (re-lubrication intervals, oil-condition monitoring, filtration); design the sealing system (the right seal type for the environment); design the bearing housing (drains, vents, drain plugs).

\subsection{Misalignment and overload}

A bearing operated outside its alignment specification (typical limit: $0.5$-$1$ minute of arc for ball bearings, more for spherical) has concentrated stress at one location of the race. The concentration accelerates fatigue. A bearing operated above its rated load follows the cube-of-load life reduction ($L_{10}$ is $\propto P^{-3}$ for ball bearings); a 25\% overload halves the life.

The engineering response: align carefully (use of dial indicators or laser alignment tools), select bearings with appropriate misalignment capability, monitor for changes in vibration that indicate misalignment development.

\subsection{Smearing, false brinelling, fluting}

\textbf{Smearing.} High-speed rolling contact with insufficient lubricant produces local welding and material transfer.

\textbf{False brinelling.} Small oscillating motion of a stationary bearing (vibration during shipping or off-state) wears small dents into the raceway at the rolling-element positions.

\textbf{Fluting (electrical erosion).} Current passing through the bearing (common in motor-driven equipment with VFDs) produces electrical discharges at the rolling contact. The discharges erode the raceway in characteristic patterns. The remedy is to isolate the bearing electrically or to provide a separate current path (slip ring, ground brush, insulated bearing).

The engineering response: identify the failure mechanism from the bearing's appearance after failure, address the root cause (transport stresses, electrical isolation, lubrication regime).

\begin{principle}[Bearings tell their story when they fail]
A bearing's failure mode is diagnostic. Spalled raceways under symmetrical loading indicate normal fatigue at end-of-life. Asymmetric wear indicates misalignment. Electrical-discharge patterns indicate VFD-induced fluting. Wash-out of grease and bright wear indicate contamination. The engineering discipline is to inspect failed bearings, identify the mode, and engineer the system to prevent recurrence.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Characterise the friction of three lubricant-surface combinations]
Build or use a small tribometer (or a pin-on-disc setup). Characterise the friction of three combinations: dry steel-on-steel, grease-lubricated steel-on-steel, and one third combination of your choice (PTFE on steel, ceramic on steel, oil-lubricated steel-on-steel).

\begin{enumerate}[leftmargin=*]
  \item Specify your apparatus: normal load, sliding speed, surface preparation, lubricant volume.
  \item Measure friction coefficient at three speeds. Plot the Stribeck curve where possible.
  \item Run for at least $10^5$ cycles or $1$ hour. Observe wear-track development.
  \item Examine the wear track under magnification ($10\times$ minimum). Identify the wear mechanism.
  \item Compute the wear coefficient (Archard's law) for each combination.
\end{enumerate}

The deliverable is the apparatus description, the measurements, the wear-track photographs, and a short report (no more than four pages) on the engineering insights from comparing the three combinations.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A ball bearing has dynamic load capacity $C = 12\,\si{\kilo\newton}$ and is loaded with $P = 2\,\si{\kilo\newton}$. Compute $L_{10}$ in revolutions and in hours at $1500\,\si{rpm}$.
\end{exercise}

\begin{exercise}
A bearing's load increases from $2$ to $3\,\si{\kilo\newton}$. Compute the life reduction.
\end{exercise}

\begin{exercise}
For a journal bearing with $\eta = 0.05\,\si{\pascal\second}$, $U = 5\,\si{\meter\per\second}$, $L = 50\,\si{\milli\meter}$, $P = 1\,\si{\mega\pascal}$, estimate the central film thickness.
\end{exercise}

\begin{exercise}
A bearing operates at $60\,\si{\degreeCelsius}$ with a $0.32\,\si{\pascal\second}$ oil at that temperature. Compute the operating viscosity if the oil temperature rises to $90\,\si{\degreeCelsius}$ (assume Walther's equation or a simple exponential drop).
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive Amontons' friction law from the asperity-contact model.
\end{exercise}

\begin{exercise}
Derive the Reynolds equation for an infinite-width thin film. State the assumptions.
\end{exercise}

\begin{exercise}
Show that the cube-of-load life reduction for ball bearings follows from a fatigue analysis with a Weibull distribution and a cube-of-stress Hertzian relation.
\end{exercise}

\begin{exercise}
Derive Archard's wear law: $V = k \cdot F \cdot s / H$, with $V$ wear volume, $F$ load, $s$ sliding distance, $H$ hardness.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the friction loss in a $1\,\si{\meter}$-diameter, $10\,\si{rpm}$ industrial-scale rotating drum supported on hydrodynamic journal bearings.
\end{exercise}

\begin{exercise}
Estimate the energy lost to bearing friction in a $50\,\si{\kilo\watt}$ electric motor running 24/7 for a year.
\end{exercise}

\begin{exercise}
Estimate the contamination-induced life reduction for a bearing operating in a workshop with airborne particulate at $10\,\si{\micro\gram\per\meter\cubed}$.
\end{exercise}

\begin{exercise}
Estimate the relubrication interval for a grease-lubricated bearing operating at $60\,\si{\degreeCelsius}$ and $1500\,\si{rpm}$.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate the Stribeck curve for a journal bearing as a function of $\eta U / P$. Identify the regime boundaries.
\end{exercise}

\begin{exercise}
Simulate the load distribution in a deep-groove ball bearing under pure radial load. Verify that the loaded zone is approximately $120°$.
\end{exercise}

\begin{exercise}
Simulate a transient bearing start-up: from rest to operating speed. Observe the transition from boundary to hydrodynamic lubrication.
\end{exercise}

\begin{exercise}
Simulate the wear-coefficient measurement (pin-on-disc) over $10^5$ cycles. Plot wear-track depth.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design the bearing set for a $5\,\si{\kilo\watt}$ electric motor with two-pole speed and an output shaft loaded at $500\,\si{\newton}$ axial, $1\,\si{\kilo\newton}$ radial.
\end{exercise}

\begin{exercise}
Design the seal arrangement for a gearbox operating in an outdoor environment with periodic rain exposure.
\end{exercise}

\begin{exercise}
Design the lubrication system for a long-life ($L_{10} > 10$ years) bearing in a wind-turbine yaw drive.
\end{exercise}

\begin{exercise}
Design the bearing-protection strategy for a variable-frequency-drive-fed motor (anti-fluting measures).
\end{exercise}

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A team's bearing fails consistently at $30\%$ of $L_{10}$. State three engineering hypotheses for the discrepancy.
\end{exercise}

\begin{exercise}
A team uses sealed-for-life bearings in a high-temperature environment. State the engineering risks.
\end{exercise}

\begin{exercise}
A team is choosing between a hydrodynamic and a rolling-element bearing for a precision spindle. State the engineering considerations.
\end{exercise}

\begin{exercise}
A team's gearbox shows accelerated bearing wear on the high-speed shaft only. State the engineering analysis.
\end{exercise}

\begin{exercise}
A team's bearing-temperature monitoring shows steady increase over weeks. State three engineering responses.
\end{exercise}

\begin{exercise}
A team replaces grease lubrication with synthetic oil to extend life. State the engineering trade-offs.
\end{exercise}

\begin{exercise}
A senior engineer argues that ``ball bearings are obsolete for high-precision applications.'' Critique. State the cases where rolling-element bearings remain preferable.
\end{exercise}

\begin{exercise}
A team's $L_{10}$ calculation gives $20$ years; field experience shows actual failures at $5$ years. State the engineering analysis and the likely cause.
\end{exercise}

\begin{exercise}
A team's vibration-monitoring shows a developing high-frequency band at the ball-pass frequency of the outer race. State the engineering diagnosis and the recommended action.
\end{exercise}

\begin{exercise}
A team is choosing between a lip seal and a labyrinth seal for a high-speed turbocharger shaft. State the engineering considerations and the trade-offs.
\end{exercise}
