\chapter{Heat engines and refrigerators}
\label{vol04:ch03}

\epigraph{The motive power of heat is independent of the agents
employed to realise it; its quantity is fixed solely by the temperatures
of the bodies between which is effected, finally, the transfer of the
caloric.}{Sadi Carnot, \emph{Reflections on the Motive Power of Fire}
(1824), in the Mendoza translation
\cite[p.~20]{hist:carnot1824}.}

\chapmeta{Half-life: foundational. Archetypes: optimisation, scaling. Exercise target: 31. Project track: estimate household car-engine efficiency three ways. Hazard class: low (paperwork, optionally a vehicle's recorded fuel-economy data). Reader path default: standard, with sections explicitly tagged. Named cases: none required in this chapter; second-law-violating efficiency claims are described by mechanism.}

\noindent
A heat engine takes thermal energy from a hot reservoir, converts part
of it to work, and rejects the remainder to a cold reservoir
(figure \ref{fig:vol04ch03:engine-refrigerator-flow}a). A
refrigerator runs the same machinery in reverse: input work moves
thermal energy from a cold reservoir to a hot one
(figure \ref{fig:vol04ch03:engine-refrigerator-flow}b). The Carnot bound
of chapter~1 fixes the maximum efficiency for given reservoir
temperatures; the working cycles of this chapter approximate the bound
by different geometries optimised for different applications. Rankine
for power plants, Brayton for jet engines and gas turbines, Otto for
spark-ignition cars, Diesel for compression-ignition engines and
ships, vapour-compression for refrigeration and air conditioning, and
combined cycles for the highest practical efficiencies.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-engine-refrigerator-flow.tex}

\section{Carnot cycle as the reference}\pathtag{core}

The \engterm{Carnot cycle} consists of four reversible processes
operating between two thermal reservoirs at $T_h$ and $T_c$
\cite{hist:carnot1824}:

\begin{enumerate}[leftmargin=*]
  \item Isothermal expansion at $T_h$: the working fluid absorbs heat
    $Q_h$ from the hot reservoir while delivering work.
  \item Adiabatic expansion from $T_h$ to $T_c$: the fluid cools as
    it continues to deliver work, with no heat transfer.
  \item Isothermal compression at $T_c$: work is done on the fluid as
    it rejects heat $Q_c$ to the cold reservoir.
  \item Adiabatic compression from $T_c$ to $T_h$: work is done on
    the fluid as it heats back to the hot-reservoir temperature.
\end{enumerate}

On the $T$-$s$ diagram the Carnot cycle is a rectangle: two horizontal
isotherms and two vertical adiabats (figure
\ref{fig:vol04ch03:carnot-ts}). The area enclosed by the
rectangle equals the net work per cycle; the area below the upper
isotherm equals the heat absorbed; the area below the lower isotherm
equals the heat rejected. The thermal efficiency is

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-carnot-ts.tex}
\[
\eta_{\text{Carnot}} = 1 - \frac{T_c}{T_h},
\]
the famous Carnot bound. The bound is universal: no cyclic device
operating between the same two reservoirs can exceed it, regardless
of the working fluid, the cycle geometry, or the engineering of the
components. The full proof is in chapter~1 of this volume; the bound is
the engineering content of the second law applied to power
production.

The universality deserves the argument in full, because it is the
argument that licenses every efficiency comparison in the rest of the
chapter. Suppose an engine $X$ existed with an efficiency above the
Carnot value between the same two reservoirs. Couple it to a reversible
Carnot engine run backward as a refrigerator between those reservoirs,
and let the work that $X$ produces drive the refrigerator
(figure \ref{fig:vol04ch03:carnot-theorem-proof}). Size the pair so the
refrigerator returns to the hot reservoir exactly the heat $Q_h$ that
$X$ drew from it. Because $X$ was assumed more efficient, it produces
more work per unit of $Q_h$ than a Carnot engine would, so it produces
more work than the Carnot refrigerator needs to pump $Q_h$ back up. The
surplus work is free for the taking, and the combined device, drawing
its heat from the cold reservoir, has moved heat from cold to hot with
no net work supplied from outside. That is precisely the self-acting
cold-to-hot heat flow the Clausius statement of the second law forbids,
so the premise fails: no engine can beat the Carnot value. The same
construction, run with two reversible engines, shows that all reversible
engines between the same reservoirs share one efficiency, since either
beating the other would let the pair violate Clausius again. The Carnot
efficiency is therefore not a property of any particular machine; it is
a property of the two temperatures alone, which is why the result
survives every later change of working fluid and cycle.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-carnot-theorem-proof.tex}

The argument also fixes the temperature scale itself. Two reversible
engines staged between $T_h$ and $T_c$ through an intermediate $T_m$ must
compose so that the heat ratios multiply, $Q_h/Q_c = (Q_h/Q_m)(Q_m/Q_c)$,
and the only function of two temperatures whose values compose this way
is a ratio of a single-argument function, $Q_h/Q_c = f(T_h)/f(T_c)$. The
choice $f(T) = T$ defines the thermodynamic, or absolute, temperature
scale, on which the reversible heat ratio equals the temperature ratio
and the Carnot efficiency takes the form $1 - T_c/T_h$ with $T$ in
kelvin. The kelvin scale of the steam tables and the gas laws is the same
scale this argument constructs, which is the reason a Carnot calculation
done in kelvin agrees with a property-table calculation done from
measured enthalpies and entropies.

The Carnot cycle is not used in practice because the four processes
require infinitely slow operation to remain reversible, and the
required working-fluid behaviour (isothermal heat transfer with the
reservoirs) places impractical demands on the heat exchangers. The
cycle is the reference against which practical cycles are measured.
A real engine operating between $T_h$ and $T_c$ achieves some
fraction of the Carnot efficiency:
\[
\eta_{\text{actual}} = \eta_{\text{Carnot}} \times \eta_{\text{second-law}},
\]
with the second-law efficiency typically $30\,\%$ to $60\,\%$ for
working engines. The gap between Carnot and actual is the working
field of cycle optimisation.

\engterm{The reversed Carnot cycle for refrigeration}. Running the
Carnot cycle in reverse (counter-clockwise on the $T$-$s$ diagram)
gives a reversible refrigerator. The coefficient of performance for
cooling is
\[
\text{COP}_{\text{Carnot},R} = \frac{T_c}{T_h - T_c},
\]
and for heating (heat pump),
\[
\text{COP}_{\text{Carnot},HP} = \frac{T_h}{T_h - T_c} =
\text{COP}_{\text{Carnot},R} + 1.
\]
Both diverge as $T_h - T_c \to 0$ and approach unity from above as
$T_h - T_c$ grows (figure \ref{fig:vol04ch03:cop-vs-temperature-lift}).
The Carnot COP is the working bound for the
performance of any refrigerator or heat pump between the same two
reservoir temperatures.

The bound on the reversed cycle follows from the same Clausius argument,
run the other way. A refrigerator claiming a coefficient of performance
above the Carnot value would, coupled to a Carnot heat engine driving it,
move heat from cold to hot while returning work to spare, again the
forbidden self-acting transfer. So $\text{COP}_R \le T_c/(T_h - T_c)$ for
every refrigerator, with equality only for a reversible one. The heat-pump
bound is not an independent fact but the same statement read for the heat
delivered: since $Q_h = Q_c + W$ for any device, dividing by the work
gives $\text{COP}_{HP} = \text{COP}_R + 1$ exactly, reversible or not, so
the heat-pump bound $T_h/(T_h - T_c)$ exceeds the refrigerator bound by
precisely one. Two readings of this identity carry into design. First, a
heat pump can never deliver less than one unit of heat per unit of work,
because the $+1$ is the work itself dissipating into the warm space, which
is why a heat pump degraded to $\text{COP}_{HP} = 1$ has simply become a
resistance heater. Second, both bounds blow up as the lift $T_h - T_c$
shrinks toward zero, so the design move with the most effect is always
to cut the temperature lift: a refrigerator or heat pump pays for every kelvin of
lift it does not need, and the heat-exchanger sizing that narrows the pinch
between the working fluid and each reservoir buys COP more cheaply than any
improvement to the compressor. The deep-lift penalty is steep enough that
halving the lift roughly doubles the bound, which is the quantitative form
of the rule that the cheapest cold is the cold you do not have to reach
far for.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-cop-vs-temperature-lift.tex}

\begin{mastery}
The $T$-$s$ diagram is not only a picture of the cycle; it is an exact
accounting of work and of work destroyed, and reading it that way turns
the second law into a measuring tool. For any reversible process the heat
transferred is the area under the path on the $T$-$s$ plane, $q = \int T
\, ds$, so the area enclosed by a reversible cycle is the net heat, which
by the first law over a cycle equals the net work. The Carnot rectangle
makes this literal: its height is the temperature lift, its width is the
entropy span $\Delta s = q_h/T_h$ of the isothermal heat intake, and its
area $(T_h - T_c)\Delta s$ is exactly $w_{\text{net}}$. The heat rejected
is the strip below the lower isotherm, $T_c \Delta s$, and the ratio of
that strip to the full column under the upper isotherm is $T_c/T_h$, the
Carnot result read straight off the geometry.

The same diagram measures irreversibility. When heat $q$ crosses a finite
temperature gap from a source at $T_1$ to a sink at $T_2 < T_1$, the
source loses entropy $q/T_1$ and the sink gains $q/T_2$, so the universe
gains $s_{\text{gen}} = q(1/T_2 - 1/T_1) > 0$. The work that transfer
forfeited, the area on the $T$-$s$ plane between where the heat could have
been delivered and where it actually was, is the lost work
\[
w_{\text{lost}} = T_0 \, s_{\text{gen}},
\]
the Gouy-Stodola relation, with $T_0$ the dead-state temperature. Every
irreversible step in a real cycle, the finite-temperature heat transfer
in a boiler, the throttle in a refrigerator, the non-isentropic
expansion in a turbine, spreads the cycle's path out in entropy and so
opens an area of lost work that a $T$-$s$ plot displays directly. The
practical reading: to find where a cycle is bleeding work, look for the
processes that move it horizontally on the $T$-$s$ plane at a temperature
far from where the heat is wanted, and attack those, because their area
is the work the second law says was thrown away.
\end{mastery}

\begin{archetype}[Optimisation under second-law constraint]
A cycle's design problem is to maximise the work output per unit heat
input, subject to the constraint that the entropy generated per cycle
be non-negative. The Carnot cycle saturates the constraint at zero;
real cycles generate entropy through finite-temperature heat
transfer, throttling, friction, and combustion at finite affinity.
The engineering art is to identify which irreversibility costs the
most, redesign the component that produces it, and reapproach the
Carnot bound. Across the chapter, the optimisation archetype recurs:
each cycle of sections 3.2 to 3.6 is a different geometric
compromise between Carnot ideality and practical heat-exchanger,
turbine, and compressor design.
\end{archetype}

\begin{mastery}
The Carnot efficiency is the bound for a reversible engine, which
requires infinitely slow operation and so produces zero power. An
engine that produces useful power must transfer heat across a finite
temperature difference, and a more useful bound asks for the efficiency
at maximum power output rather than the efficiency at zero power. For an
engine with finite-conductance heat exchangers on both reservoirs, the
efficiency at maximum power is the Curzon-Ahlborn value
\[
\eta_{\text{CA}} = 1 - \sqrt{\frac{T_c}{T_h}},
\]
which lies below the Carnot value $1 - T_c/T_h$. For a $560\,\degC$
boiler and a $30\,\degC$ condenser, the Carnot bound is $64\,\%$ while
the Curzon-Ahlborn value is $1 - \sqrt{303/833} = 40\,\%$, close to the
efficiency that real plants achieve. The agreement is partly
coincidental (real plants are not optimised for maximum power, and their
losses are not the single finite-conductance loss the model assumes),
but the lesson holds: the reversible bound overstates what a
power-producing machine can reach, and a finite-rate bound is the right
yardstick for a machine that must deliver power, not just work.
\end{mastery}

\begin{mastery}
The Curzon-Ahlborn result assumed a fixed conductance on each reservoir; a
designer given a fixed total to spend must also decide how to split it between the
hot and cold heat exchangers, and minimising the entropy the two of them generate
answers that. Model the engine as reversible inside, with all irreversibility in
the two finite-conductance exchangers: the hot one passes $Q_h$ across a
conductance $U_h A_h$ and the cold one passes $Q_c$ across $U_c A_c$, subject to a
fixed total area $A_h + A_c = A$. Each exchanger generates entropy at a rate set by
its heat flow and its temperature drop, and for a target power the total entropy
generation, $\dot S_{\text{gen}} = Q_c/T_c - Q_h/T_h$ read at the reservoir
temperatures, is minimised when the area is allocated so that the two exchangers
carry equal conductance-weighted duty. Carrying the Lagrange condition through, the
optimum splits the area in the ratio
\[
\frac{A_h}{A_c} = \sqrt{\frac{U_c\,Q_h}{U_h\,Q_c}},
\]
so with equal heat-transfer coefficients on the two sides the areas divide as the
square root of the heat-flow ratio, giving the hot exchanger the larger share
because it must pass the larger heat $Q_h = Q_c + W$. Two readings carry into
design. The optimum is a balance, not a corner: starving either exchanger drives
its temperature drop up and its entropy generation with it, so a design that
lavishes area on the boiler and skimps on the condenser wastes the same work
potential as the reverse. And the allocation rule is why real plants size the
condenser generously despite its lower temperature: the cold exchanger rejects
nearly as much heat as the hot one absorbs, so it needs comparable area, and a
cramped condenser is a common cause of a plant falling short of its rated
efficiency. Entropy-generation minimisation, applied component by component, turns
the second law from a bound into a sizing tool, the same move the Gouy-Stodola
relation made for diagnosis.
\end{mastery}

\section{Rankine cycle}\pathtag{core}

The \engterm{Rankine cycle} is the working power-plant cycle: it
generates more than half of the world's electricity through coal,
natural-gas combined cycles (where the Rankine cycle is the bottom
half), nuclear, geothermal, and concentrated-solar plants. The
working fluid is water; the cycle has four components and four
states.

\begin{enumerate}[leftmargin=*]
  \item Pump (1 to 2): liquid water is pressurised from condenser
    pressure $p_c$ to boiler pressure $p_b$. Specific work input
    $w_p = v_f(p_b - p_c)$, small compared to the turbine work.
  \item Boiler (2 to 3): the high-pressure water is heated to
    saturated or superheated vapour at $p_b$. Specific heat input
    $q_h = h_3 - h_2$.
  \item Turbine (3 to 4): the vapour expands to condenser pressure
    $p_c$, delivering work $w_t = h_3 - h_4$. The expansion is
    nearly isentropic in a well-designed turbine.
  \item Condenser (4 to 1): the low-pressure two-phase mixture
    condenses to saturated liquid at $p_c$, rejecting heat $q_c =
    h_4 - h_1$.
\end{enumerate}

The thermal efficiency is
\[
\eta_R = \frac{w_t - w_p}{q_h} = 1 - \frac{q_c}{q_h}.
\]

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-rankine-ts.tex}

On a $T$-$s$ diagram (figure \ref{fig:vol04ch03:rankine-ts}), the cycle
is bounded by two isobars (the boiler
and condenser pressures) and the turbine and pump processes. The
boiler isobar dips into the two-phase region and rises into the
superheated regime; the condenser isobar lies in the two-phase region
at low temperature. The cycle's efficiency depends strongly on the
boiler and condenser conditions:

\begin{itemize}[leftmargin=*]
  \item Raising the boiler pressure raises the average temperature
    of heat addition and increases efficiency. Modern supercritical
    plants operate at boiler pressures of $25$ to $30\,\text{MPa}$.
  \item Raising the boiler temperature (superheat, reheat) raises the
    average temperature of heat addition further and increases
    efficiency.
  \item Lowering the condenser pressure (and hence the condenser
    saturation temperature) lowers the average temperature of heat
    rejection and increases efficiency. Practical plants run with
    condensers at $4$ to $10\,\text{kPa}$, with the saturation
    temperature limited by the cooling-water supply.
\end{itemize}

\engterm{Reheat Rankine cycle}. After partial expansion through the
high-pressure turbine, the steam returns to the boiler for reheating
back to a high temperature, then completes its expansion in the low-
pressure turbine. Reheat raises the average heat-addition temperature
and (more importantly) raises the steam quality at the low-pressure
turbine exit, limiting blade erosion. Most large power plants use
single or double reheat.

\engterm{Regenerative Rankine cycle}. Some steam is bled from the
turbine at intermediate pressures and used to heat the feedwater
before the boiler. The bled-steam regenerative feedwater heating
raises the average temperature of heat addition and reduces the
heat that must be supplied at the boiler from the high-temperature
flame. Practical large plants use six to ten regenerative
feedwater heaters.

Modern coal-fired Rankine plants achieve net thermal efficiencies of
$40$ to $45\,\%$; ultra-supercritical plants reach $47\,\%$.
Nuclear Rankine plants are limited to lower turbine inlet
temperatures (typically $280\,\degC$ saturated steam) for fuel-cladding
reasons and achieve efficiencies near $33\,\%$. The Carnot bound
between a $300\,\degC$ boiler and a $30\,\degC$ condenser is
$1 - 303/573 = 47\,\%$; the Carnot bound between a $560\,\degC$
boiler and a $30\,\degC$ condenser is $1 - 303/833 = 64\,\%$. Working
Rankine plants reach the Carnot bound to within a factor of $0.6$ to
$0.7$ \cite{text:moran-shapiro-thermo}.

\subsection{Turbine-exit wetness and why it limits the cycle}

The efficiency levers on the Rankine cycle all push in one direction, toward a
hotter, higher-pressure boiler and a colder, lower-pressure condenser, and left
alone they would drive the turbine expansion deep into the two-phase region, where
the exhaust is a mist of water droplets in steam. That wetness is the constraint
that quietly sets how far the levers may be pushed. Droplets condensing in the
last turbine stages travel much slower than the vapour carrying them, so they
strike the moving blades on their backs at a high relative velocity and erode the
blade leading edges, a slow mechanical attrition that thins and pits the metal
over years of running. The trailing stages of a large steam turbine, where the
volume flow is greatest and the blades are longest and most highly stressed, are
exactly where the steam is wettest, so erosion and mechanical load compound. The
practical rule that has governed steam-plant design for a century is to keep the
turbine-exit quality above about $0.88$, which is the number the reheat cycle
exists to protect: the single-expansion cycle of the worked example sat at
$0.82$, uncomfortably wet, while adding a reheat lifted the exit to $0.92$,
comfortably dry. Wetness, not efficiency, is the first reason large plants reheat,
and it is why a plant chasing efficiency by raising boiler pressure must reheat to
keep its last stages out of the erosion region.

\begin{mastery}
Wetness costs efficiency as well as blades, and the loss follows a simple linear
rule. Water droplets suspended in the expanding steam do not contribute their
share of work to the blades, and the momentum spent flinging them out of the flow
path is work removed from the shaft, so a turbine stage running wet is less
efficient than the same stage running dry. The \engterm{Baumann rule} states that
each one percent of average wetness through a stage costs roughly one percent of
that stage's isentropic efficiency,
\[
\eta_{\text{wet}} = \eta_{\text{dry}}\left(1 - \alpha\,\bar{y}\right),
\quad \alpha \approx 1,
\]
with $\bar{y}$ the mean moisture fraction (one minus the mean quality) across the
stage. A low-pressure stage expanding from quality $0.95$ to $0.88$ carries a mean
moisture near $0.085$, so it loses about eight and a half percent of its
isentropic efficiency to wetness alone, on top of the profile and leakage losses
every stage carries. The rule explains two design moves at once. It is a second
reason to reheat, since lifting the exit quality cuts $\bar{y}$ and recovers the
Baumann loss across the whole low-pressure turbine, not just the last stage; and
it is the reason large machines fit moisture-removal grooves and drainage belts in
the late stages, bleeding off the film of water that collects on the casing before
it re-entrains into the flow. The Baumann coefficient $\alpha$ is not exactly one
for every machine, but taking it as one gives the right size of the penalty and
the right sign of every design response to it.
\end{mastery}

\section{Brayton cycle}\pathtag{core}

The \engterm{Brayton cycle} is the working cycle of jet engines and
land-based gas turbines. The working fluid is air (drawn from
atmosphere, mixed with fuel and combusted, and exhausted back to
atmosphere) or, in a closed Brayton cycle, helium or carbon dioxide.
The four processes:

\begin{enumerate}[leftmargin=*]
  \item Compressor (1 to 2): air is compressed adiabatically from
    atmospheric to combustor inlet pressure. Pressure ratio $r_p =
    p_2/p_1$.
  \item Combustor (2 to 3): fuel is burned in the compressed air at
    approximately constant pressure, raising the temperature to the
    turbine inlet temperature $T_3$.
  \item Turbine (3 to 4): the high-pressure hot gas expands
    adiabatically back to atmospheric pressure, delivering work.
  \item Exhaust (4 to 1): the gas is rejected to atmosphere (open
    cycle) or cooled in a heat exchanger and returned to the
    compressor (closed cycle).
\end{enumerate}

The ideal-gas air-standard Brayton cycle's thermal efficiency depends
only on the pressure ratio:
\[
\eta_B = 1 - \frac{1}{r_p^{(\gamma-1)/\gamma}}.
\]
For air with $\gamma = 1.4$, the efficiency rises from $37\,\%$ at
$r_p = 5$ to $54\,\%$ at $r_p = 15$ to $62\,\%$ at $r_p = 30$ to $67\,\%$
at $r_p = 50$. Modern aero-derivative gas turbines run at pressure
ratios above $40$; large industrial gas turbines run at $15$ to $20$.
The companion \texttt{cycle\_efficiencies.py} reproduces these figures and
the Otto and Diesel curves, and \texttt{brayton\_cycle.py} returns the four
state temperatures and the back-work ratio for any inlet condition.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-brayton-pv-ts.tex}

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-efficiency-vs-ratio.tex}

Figure \ref{fig:vol04ch03:brayton-pv-ts} traces the cycle on both the
$p$-$v$ and $T$-$s$ planes, and figure \ref{fig:vol04ch03:efficiency-vs-ratio}
plots the Otto and Brayton efficiency curves against ratio so the
diminishing return on each added unit of ratio is visible at a glance.

The efficiency formula understates the practical bound. The Brayton
cycle is limited by the maximum allowable turbine inlet temperature,
set by the metallurgy of the turbine blades and the effectiveness of
blade cooling. Modern aero-engines run with $T_3$ above $1700\,
\si{\kelvin}$, well above the un-cooled blade material's melting
point; the blades survive through internal air-channel cooling, film
cooling, and ceramic thermal-barrier coatings. The cycle's actual
efficiency depends on $r_p$ and $T_3$ together; for each $T_3$ there
is an optimum $r_p$ that maximises specific work, and a higher $r_p$
that maximises thermal efficiency.

The split between those two optima is the recurring tension of
gas-turbine design. Specific work, the net work per kilogram of air, sets
how large and heavy a machine must be to make a given power, and it peaks
at a moderate pressure ratio because at low $r_p$ the cycle does little
work while at high $r_p$ the compressor consumes most of what the turbine
makes. Thermal efficiency, by the air-standard formula, climbs
monotonically with $r_p$, so it wants a higher ratio than specific work
does. An aero-engine, where weight is the master constraint, leans toward
the specific-work optimum and accepts a lower efficiency; a stationary
power turbine, where fuel cost dominates over a long life, leans toward
the efficiency optimum and accepts a heavier machine. Raising $T_3$ shifts
both optima to higher $r_p$ and raises the specific work at every ratio,
which is why the relentless climb in turbine inlet temperature, sustained
by blade cooling and thermal-barrier coatings, has driven gas-turbine
performance for decades. Each $100\,\si{\kelvin}$ of additional inlet
temperature buys both more work per kilogram and headroom to raise the
pressure ratio, and the metallurgy of the first turbine stage is the
binding constraint on how far the trade can be pushed.

\engterm{Regenerative Brayton cycle}. Exhaust heat from the turbine
exit (still at several hundred Celsius) is recovered to preheat the
combustor inlet air. The regenerator (a counter-flow heat exchanger)
recovers some of the rejected heat and reduces the heat required in
the combustor. Ground-based gas turbines for cogeneration or
combined-cycle use frequently include regeneration; aero-engines do
not, because the regenerator's weight is prohibitive.

\engterm{Intercooled and reheated Brayton}. Compression in two stages
with cooling between stages reduces compressor work; expansion in
two stages with reheating between stages increases turbine work. The
combined intercooled-regenerative-reheat Brayton cycle approaches a
Carnot cycle in the limit of many intercooler and reheat stages.
Practical implementations use one or two of each.

\subsection{The specific-work optimum and where it sits relative to peak efficiency}

The split between the pressure ratio that maximises efficiency and the one
that maximises work per kilogram is not a vague tendency; it has a closed
form, and deriving it shows why the two optima can never coincide. Write
the temperature ratio of the cycle as $\tau = T_3/T_1$, the ratio of
turbine inlet to compressor inlet, and the isentropic pressure-ratio factor
as $x = r_p^{(\gamma-1)/\gamma}$, so that the compressor exit is $T_2 = T_1
x$ and the isentropic turbine exit is $T_4 = T_3/x = T_1 \tau/x$. The net
specific work, with constant $c_p$, is the turbine work less the compressor
work,
\[
\frac{w_{\text{net}}}{c_p T_1} = \tau\!\left(1 - \frac{1}{x}\right) -
(x - 1),
\]
a function of the single variable $x$ for a fixed $\tau$. Differentiating
and setting the derivative to zero,
\[
\frac{d}{dx}\!\left[\tau\!\left(1 - \frac{1}{x}\right) - (x - 1)\right]
= \frac{\tau}{x^2} - 1 = 0
\quad\Longrightarrow\quad x^\star = \sqrt{\tau},
\]
so the specific work peaks where the pressure-ratio factor equals the
square root of the temperature ratio. In terms of the pressure ratio
itself,
\[
r_p^\star = \tau^{\gamma/[2(\gamma-1)]},
\]
the geometric mean of unity and the ceiling ratio $\tau^{\gamma/(\gamma-1)}$
at which the compressor exit reaches the turbine inlet temperature and no
heat can be added (figure \ref{fig:vol04ch03:brayton-specific-work}). At
that optimum the compressor exit and the isentropic turbine exit sit at the
same temperature, $T_2 = T_4 = T_1\sqrt{\tau}$, a tidy fact with a
consequence for regeneration the next subsection turns on.

The thermal efficiency, by contrast, is $\eta = 1 - 1/x$, which climbs
monotonically in $x$ and so has no interior maximum: it wants the largest
pressure ratio the ceiling allows. The two optima therefore split by
construction, and the gap widens with $\tau$. For $\tau = 4.67$ ($T_3 =
1400\,\si{\kelvin}$, $T_1 = 300\,\si{\kelvin}$) the specific-work optimum is
$x^\star = 2.16$, that is $r_p^\star = 14.8$, while the efficiency keeps
rising past $r_p = 30$. An aero-engine sized by weight sits near
$r_p^\star$; a stationary machine sized by fuel bill sits well above it.
Raising $\tau$ pushes $r_p^\star$ up as $\tau^{1.75}$ and raises the peak
specific work, the quantitative engine behind the decades-long climb in
turbine inlet temperature.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-brayton-specific-work.tex}

\subsection{Regeneration effectiveness and the efficiency it buys}

A regenerator transfers heat from the turbine exhaust to the compressed air
before the combustor, and its benefit is bounded by how completely it makes
that transfer. The \engterm{regenerator effectiveness} $\varepsilon$ is the
ratio of the actual enthalpy rise of the cold stream to the maximum the
exhaust temperature would permit,
\[
\varepsilon = \frac{T_x - T_2}{T_4 - T_2},
\]
where $T_x$ is the temperature the compressed air reaches leaving the
regenerator, $T_2$ the compressor exit, and $T_4$ the turbine exit. An ideal
regenerator ($\varepsilon = 1$) heats the air all the way to $T_4$; a real
counter-flow unit reaches $\varepsilon = 0.7$ to $0.9$, limited by its
surface area and the pressure drop a larger matrix imposes. With
regeneration the combustor need only supply $q_h = c_p(T_3 - T_x)$ instead
of $c_p(T_3 - T_2)$, while the net work is unchanged, so the efficiency
rises. For the ideal case the heat added is $c_p(T_3 - T_4)$ and the
efficiency reduces to
\[
\eta_{\text{reg}} = 1 - \frac{T_1}{T_3}\,r_p^{(\gamma-1)/\gamma}
= 1 - \frac{x}{\tau},
\]
a result that falls with pressure ratio, the opposite slope to the simple
cycle's $1 - 1/x$. The two expressions are equal where $x = \sqrt{\tau}$,
exactly the specific-work optimum, because there $T_2 = T_4$ and the
regenerator has no temperature gap to feed on (figure
\ref{fig:vol04ch03:regeneration-efficiency}). Below that crossover the
turbine exit is hotter than the compressor exit and the regenerator pays;
above it the compressor exit is hotter and a regenerator would carry heat
the wrong way, so it must be bypassed. A partial regenerator interpolates
between the two limits. With the regenerator delivering air at $T_x = T_2 +
\varepsilon(T_4 - T_2)$, the heat the combustor must supply is $q_h =
c_p(T_3 - T_x)$ while the net work stays $c_p[(T_3 - T_4) - (T_2 - T_1)]$,
so
\[
\eta = \frac{(T_3 - T_4) - (T_2 - T_1)}
{(T_3 - T_2) - \varepsilon\,(T_4 - T_2)}
= \frac{\tau(1 - 1/x) - (x - 1)}
{(\tau - x) - \varepsilon\,(\tau/x - x)},
\]
which returns the simple cycle at $\varepsilon = 0$ and the ideal
regenerator $1 - x/\tau$ at $\varepsilon = 1$. The efficiency gain over the
simple cycle scales linearly in $\varepsilon$ for a fixed pressure ratio, so
a regenerator of effectiveness $0.8$ captures four-fifths of the lift an
ideal one would. The design reading is that regeneration is a
low-pressure-ratio refinement: it lifts the efficiency of a modest-ratio
machine substantially, does nothing at the crossover, and is pointless above
it, which is why heavy industrial machines for cogeneration adopt it and
high-ratio aero-derivatives do not.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-regeneration-efficiency.tex}

\subsection{The optimum intermediate pressure for intercooling and reheat}

Splitting the compression into two stages with a cooler between them poses a
small optimisation: where to set the intermediate pressure. Take a compressor
lifting from $p_1$ to $p_2$ in two stages through an intermediate $p_i$, with
the gas cooled back to the inlet temperature $T_1$ between stages (perfect
intercooling). Each stage's work is $c_p T_1$ times its pressure-ratio
factor less one, so the total stage work is
\[
\frac{w_c}{c_p T_1} = \left[\left(\frac{p_i}{p_1}\right)^{(\gamma-1)/\gamma}
- 1\right] + \left[\left(\frac{p_2}{p_i}\right)^{(\gamma-1)/\gamma} - 1\right].
\]
Differentiating with respect to $p_i$ and setting the result to zero gives
the condition that the two stage pressure ratios be equal,
\[
\frac{p_i}{p_1} = \frac{p_2}{p_i}
\quad\Longrightarrow\quad
p_i = \sqrt{p_1 p_2},
\]
the intermediate pressure is the geometric mean of the end pressures, and
each stage carries the same ratio $\sqrt{p_2/p_1}$. The same argument, run on
the expansion with reheat back to the turbine inlet temperature between
stages, places the reheat pressure at the geometric mean of the turbine
inlet and exit pressures and equalises the two expansion ratios. With $N$
stages the optimum sets every stage to the ratio $(p_2/p_1)^{1/N}$, and as
$N$ grows the staged compression approaches isothermal compression at $T_1$
and the staged expansion approaches isothermal expansion at $T_3$, which is
why the fully intercooled, reheated, regenerative Brayton cycle approaches
the Carnot efficiency in the limit. The geometric-mean rule is the same one
the two-stage refrigeration example used to size its flash-intercooler
pressure; it recurs wherever a multiplicative span is split to minimise a sum
of stage works.

The Brayton cycle's advantage over Rankine is its compactness and
high specific power: a Brayton turbine can produce $10\,\text{MW}/
\text{tonne}$, against $\sim 1\,\text{MW}/\text{tonne}$ for Rankine
machinery. Its disadvantage is its lower efficiency at single-cycle
operation; the combined cycle (Brayton on top, Rankine on bottom)
solves this and is the modern standard for large new power
generation.

\section{Otto, Diesel, Atkinson cycles}\pathtag{standard}

The reciprocating internal combustion engines of cars, trucks, ships,
and stationary generators use variants of three cycles, all derived
from the same piston-cylinder geometry but differing in how fuel is
introduced and combusted.

\engterm{The Otto cycle} models the spark-ignition gasoline engine.
The four strokes (intake, compression, power, exhaust) map onto a
four-process cycle with the working fluid treated as air:

\begin{enumerate}[leftmargin=*]
  \item Isentropic compression (1 to 2) from bottom dead centre to
    top dead centre, compression ratio $r = V_1/V_2$.
  \item Constant-volume heat addition (2 to 3): combustion at top
    dead centre, treated as instantaneous, raising the gas
    temperature to $T_3$.
  \item Isentropic expansion (3 to 4) back to bottom dead centre.
  \item Constant-volume heat rejection (4 to 1): exhaust and intake,
    treated as instantaneous, returning the gas to its initial
    temperature.
\end{enumerate}

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-otto-diesel-pv.tex}

The Otto cycle's air-standard thermal efficiency is
\[
\eta_O = 1 - \frac{1}{r^{\gamma-1}},
\]
depending only on the compression ratio. For $\gamma = 1.4$ and $r =
8$ to $10$ (typical for modern gasoline engines), $\eta_O$ ranges
from $56\,\%$ to $60\,\%$. The real engine's brake thermal efficiency
is roughly $25\,\%$ to $35\,\%$ at the design operating point, with
the gap to the Otto ideal due to: combustion at non-constant volume,
heat loss to cylinder walls during combustion and expansion, friction
in the engine, pumping losses, incomplete combustion, finite valve-
opening dynamics, and exhaust gas residual.

The Otto compression ratio is limited by the fuel's auto-ignition
behaviour. A gasoline-air mixture compressed to too high a ratio will
self-ignite during the compression stroke, producing knock, which
damages the engine. The fuel's octane number measures its resistance
to auto-ignition; higher-octane fuels permit higher compression
ratios and higher efficiency. Variable-compression-ratio engines
adjust the geometry in response to load to operate at the highest
allowable compression at each load.

\engterm{The Diesel cycle} models the compression-ignition engine
(figure \ref{fig:vol04ch03:otto-diesel-pv}b, against the Otto cycle in
panel a). The differences from the Otto cycle:

\begin{itemize}[leftmargin=*]
  \item Compression is of air alone, not air-fuel mixture; the
    compression ratio is therefore not limited by fuel auto-ignition.
    Typical compression ratios are $14$ to $22$ for Diesel engines,
    against $8$ to $10$ for gasoline.
  \item Fuel is injected at top dead centre and burns as it is
    injected; the combustion is modelled as constant-pressure
    (process 2 to 3) rather than constant-volume.
\end{itemize}

The Diesel cycle's air-standard thermal efficiency is
\[
\eta_D = 1 - \frac{1}{r^{\gamma-1}} \cdot \frac{r_c^\gamma - 1}{
\gamma(r_c - 1)},
\]
where $r_c = V_3/V_2$ is the cut-off ratio (the volume ratio at the
end of combustion to that at top dead centre). For comparable
compression ratios, the Diesel efficiency is slightly less than the
Otto, because the constant-pressure heat addition occurs at a lower
average temperature than constant-volume. The Diesel advantage comes
from the higher compression ratio it can tolerate. Working Diesel
engines achieve brake thermal efficiencies of $40\,\%$ to $50\,\%$
(for marine engines and the most efficient automotive Diesels) and
in many applications outperform comparable gasoline engines.

\engterm{The Atkinson cycle} differs from Otto by separating the
expansion ratio from the compression ratio: the effective compression
is reduced (by late-closing or early-closing the intake valve), while
the geometric expansion stroke is the full piston travel. The
expansion ratio exceeds the effective compression ratio, recovering
more work from the combustion gas before exhaust. The thermal
efficiency rises (typically $5\,\%$ to $10\,\%$ relative to Otto) at
the cost of reduced specific power. The Atkinson cycle is the working
cycle of most hybrid-electric vehicles (Toyota Prius, Honda Insight,
many others), where the electric drive compensates for the reduced
specific power.

The mechanism is worth stating plainly, because it shows where the extra
work comes from. In an Otto cycle the gas at the end of expansion is still
well above atmospheric pressure when the exhaust valve opens; the energy
in that residual pressure is dumped to the exhaust, having done no work.
The Atkinson arrangement lets the gas expand further, against the longer
expansion stroke, so it leaves the cylinder closer to atmospheric
pressure and surrenders less unused pressure energy. The price is that
the same long stroke is not filled on the intake, so each cycle processes
less air and makes less power for a given engine size. The
\engterm{Miller cycle} reaches the same expansion-greater-than-compression
geometry through valve timing combined with forced induction, so that a
supercharger or turbocharger restores the lost charge mass; it is the
Atkinson idea adapted to engines that cannot afford the specific-power
penalty. Both cycles trade peak power for efficiency, and both make sense
exactly where a second power source, an electric motor or a boost system,
covers the demand peaks the downsized air path can no longer meet.

\subsection{Mean effective pressure and the Atkinson efficiency}

The thermal efficiency of a reciprocating cycle says what fraction of the
heat becomes work; the \engterm{mean effective pressure} says how much work a
given engine size produces, and it is the number that sets displacement. It
is defined as the net work per cycle divided by the displaced volume, the
constant pressure that, acting through the swept volume, would do the same
work,
\[
\text{MEP} = \frac{w_{\text{net}}}{v_1 - v_2}
= \frac{w_{\text{net}}}{v_1\,(1 - 1/r)},
\]
with $r = v_1/v_2$ the compression ratio. Two engines of equal efficiency
but different mean effective pressure differ in size for the same power: the
higher-MEP engine is smaller, which is the whole point of supercharging,
since forcing more charge mass into the same cylinder raises $w_{\text{net}}$
and so the MEP. For the Otto cycle, with $w_{\text{net}} = \eta_O
q_{\text{in}}$ and $v_1 = RT_1/p_1$, the mean effective pressure follows in
closed form,
\[
\text{MEP} = \frac{\eta_O\, q_{\text{in}}}{(RT_1/p_1)(1 - 1/r)}
= \frac{p_1\, \eta_O\, q_{\text{in}}}{R T_1\,(1 - 1/r)},
\]
which the worked Otto example evaluates to $1529\,\text{kPa}$ for $r = 9$ and
$q_{\text{in}} = 2000\,\text{kJ}/\text{kg}$. The MEP rises with intake
pressure $p_1$ (the supercharging lever), with the heat added per cycle
$q_{\text{in}}$ (the fuelling lever), and weakly with compression ratio
through both $\eta_O$ and the displacement factor.

The Atkinson cycle's efficiency follows from making the expansion ratio
$r_e = v_4/v_3$ exceed the compression ratio $r_c = v_1/v_2$. Heat is added
at constant volume, $q_{\text{in}} = c_v(T_3 - T_2)$ with $T_2 = T_1
r_c^{\gamma-1}$; the gas then expands isentropically through the full ratio
$r_e$ to $T_4 = T_3 r_e^{-(\gamma-1)}$ and rejects heat to the atmosphere at
constant pressure, not constant volume, because the longer expansion returns
the gas to atmospheric pressure rather than to the starting volume. The heat
rejected is then $q_{\text{out}} = c_p(T_4 - T_1)$, carrying $c_p$ rather
than $c_v$ because the rejection is isobaric, and the efficiency is
\[
\eta_A = 1 - \frac{q_{\text{out}}}{q_{\text{in}}}
= 1 - \gamma\,\frac{T_4 - T_1}{T_3 - T_2}.
\]
The factor $\gamma$ in front of the temperature ratio is the formal trace of
the over-expansion: the longer stroke captures the residual pressure energy
that an Otto cycle dumps at exhaust-valve opening, and the isobaric rejection
that replaces the Otto isochore is what lets it. For a representative case,
$r_c = 8$, $r_e = 12$, $T_1 = 300\,\si{\kelvin}$, and the same
$q_{\text{in}}/c_v$ as the Otto example, the Atkinson efficiency lands a few
points above the Otto value at the same compression ratio, the gain the
hybrid-vehicle engines of the previous subsection harvest. The price is in
the MEP: the same combustion now drives a larger expansion volume, so the
work per unit displaced volume falls, and the engine makes less power for its
size, which is the specific-power penalty stated qualitatively above made
quantitative here.

\subsection{Otto and Diesel compared at equal peak pressure}

Comparing the two reciprocating cycles at equal compression ratio, as the
exercises do, flatters the Otto cycle, because a real engine is limited not
by compression ratio but by the peak cylinder pressure its structure can
carry. At equal peak pressure the comparison reverses, and the reversal is
the engineering reason the Diesel cycle exists. The Otto cycle adds its heat
at constant volume, so its peak pressure climbs in proportion to the heat
added; reaching a high mean effective pressure at a fixed peak-pressure
ceiling forces the Otto cycle to a low compression ratio, and its efficiency
$1 - r^{-(\gamma-1)}$ falls with it. The Diesel cycle adds its heat at
constant pressure, so its peak pressure is just the compression pressure
$p_2 = p_1 r^\gamma$, fixed by the compression ratio alone and independent of
how much fuel is burned. A Diesel engine can therefore raise its compression
ratio until the compression pressure reaches the structural ceiling, then add
as much heat as combustion allows without raising the peak further. At a
common peak pressure the Diesel runs the higher compression ratio, recovers
the efficiency the constant-pressure heat addition costs it at equal
compression ratio, and ends ahead. The worked Diesel example with a peak-pressure
limit traces this trade with numbers; the principle is that the right
comparison between two cycles is made at the constraint that actually binds,
which for reciprocating engines is peak pressure, not compression ratio.

\begin{principle}[Compare cycles at the binding constraint]
Two cycles ranked at equal values of a parameter that does not bind in
practice can rank in the opposite order at the constraint that does. The
air-standard Otto efficiency exceeds the Diesel at equal compression ratio,
yet the Diesel wins in service because the binding constraint is peak
cylinder pressure, which the Diesel's constant-pressure heat addition holds
at the compression value while the Otto's constant-volume addition drives it
up with every increment of fuel. Identify the constraint that limits the real
machine before ranking designs by an idealised efficiency.
\end{principle}

\section{Worked examples across the cycles}\pathtag{standard}

The worked examples that follow fix the cycle arithmetic the exercises
later apply at scale. Each carries its numbers to the point where the
answer is interpretable; the table values follow the standard
thermodynamics references \cite{text:cengel-boles,
text:moran-shapiro-thermo}.

\paragraph{Otto cycle, compression ratio nine.} Take $r = 9$, inlet
$T_1 = 300\,\si{\kelvin}$ and $p_1 = 100\,\text{kPa}$, and a heat input
$q_{\text{in}} = 2000\,\text{kJ}/\text{kg}$, with cold-air-standard
constants $\gamma = 1.4$, $c_v = 0.718\,\text{kJ}/\text{kg}\cdot\text{K}$.
The isentropic compression gives $T_2 = T_1 r^{\gamma-1} =
300 \times 9^{0.4} = 722\,\si{\kelvin}$ and $p_2 = p_1 r^\gamma =
100 \times 9^{1.4} = 2167\,\text{kPa}$. Constant-volume combustion adds
$q_{\text{in}}/c_v = 2000/0.718 = 2786\,\si{\kelvin}$, so $T_3 = 722 +
2786 = 3508\,\si{\kelvin}$ and $p_3 = p_2 T_3/T_2 = 10{,}500\,\text{kPa}$.
The efficiency is $\eta_O = 1 - 9^{-0.4} = 0.585$, so the net work is
$0.585 \times 2000 = 1170\,\text{kJ}/\text{kg}$. The peak temperature
$3508\,\si{\kelvin}$ is far above any real flame temperature, which is
the first sign that the air-standard model overstates the real cycle: a
real charge dissociates and loses heat to the walls long before reaching
$3500\,\si{\kelvin}$. The solver \texttt{otto\_cycle.py} reproduces every
number here, including the mean effective pressure of $1528\,\text{kPa}$.

\paragraph{Brayton cycle, pressure ratio twelve.} Take $r_p = 12$,
$T_1 = 300\,\si{\kelvin}$, $T_3 = 1400\,\si{\kelvin}$, $\gamma = 1.4$,
$c_p = 1.005\,\text{kJ}/\text{kg}\cdot\text{K}$. The isentropic
exponent is $(\gamma-1)/\gamma = 0.286$, so $T_2 = 300 \times 12^{0.286}
= 610\,\si{\kelvin}$ and $T_4 = 1400 \times 12^{-0.286} = 688\,\si{
\kelvin}$. The turbine work is $c_p(T_3 - T_4) = 1.005 \times 712 =
715\,\text{kJ}/\text{kg}$; the compressor work is $c_p(T_2 - T_1) =
1.005 \times 310 = 312\,\text{kJ}/\text{kg}$; the net work is
$715 - 312 = 404\,\text{kJ}/\text{kg}$. The heat input is $c_p(T_3 -
T_2) = 1.005 \times 790 = 794\,\text{kJ}/\text{kg}$, so $\eta_B =
404/794 = 0.508$, matching $1 - 12^{-0.286}$. The back-work ratio
$w_c/w_t = 312/715 = 0.44$ is the Brayton cycle's defining burden:
nearly half the turbine output is consumed driving the compressor,
which is why a small drop in compressor or turbine efficiency erases a
large fraction of the net work.

\paragraph{Vapour-compression refrigerator, R-134a.} Take an
evaporator at $-10\,\degC$ and a condenser at $40\,\degC$, with an ideal
(isentropic) compressor, saturated vapour at the evaporator exit, and
saturated liquid at the condenser exit. From the saturation table in
\texttt{r134a\_saturation.csv}, the evaporator-exit vapour has $h_1 =
392.4\,\text{kJ}/\text{kg}$ and the condenser-exit liquid has $h_3 =
256.4\,\text{kJ}/\text{kg}$. Following the constant-entropy line from
state 1 up to the condenser pressure gives $h_2 \approx
425.6\,\text{kJ}/\text{kg}$. The throttle is isenthalpic, so $h_4 = h_3
= 256.4\,\text{kJ}/\text{kg}$. The cooling effect is $q_c = h_1 - h_4 =
136\,\text{kJ}/\text{kg}$; the compressor work is $w_c = h_2 - h_1 =
33.2\,\text{kJ}/\text{kg}$; the cooling COP is $136/33.2 = 4.1$. The
Carnot COP between $263\,\si{\kelvin}$ and $313\,\si{\kelvin}$ is
$263/50 = 5.3$, so the ideal cycle reaches $77\,\%$ of Carnot, and the
throttling loss (an irreversible pressure drop replacing a reversible
expander) accounts for most of the gap.

\paragraph{Reheat Rankine, where the second expansion pays off.} Take a
boiler at $15\,\text{MPa}$ and $550\,\degC$, a reheat back to $550\,\degC$
at an intermediate pressure of $2\,\text{MPa}$, and a condenser at
$10\,\text{kPa}$, with isentropic turbine and pump (figure
\ref{fig:vol04ch03:reheat-rankine-ts} traces the path). From the
superheated tables, the high-pressure turbine inlet has $h_3 = 3448\,
\text{kJ}/\text{kg}$ and $s_3 = 6.52\,\text{kJ}/\text{kg}\cdot\text{K}$.
Expanding isentropically to $2\,\text{MPa}$ lands in the superheated
region at $h_4 = 2902\,\text{kJ}/\text{kg}$, so the high-pressure stage
delivers $h_3 - h_4 = 546\,\text{kJ}/\text{kg}$. Reheating at
$2\,\text{MPa}$ back to $550\,\degC$ gives $h_5 = 3578\,\text{kJ}/\text{kg}$
and $s_5 = 7.57\,\text{kJ}/\text{kg}\cdot\text{K}$. The low-pressure
expansion to $10\,\text{kPa}$ reaches a quality $x_6 = (7.57 - 0.649)/
7.50 = 0.923$ and $h_6 = 191.8 + 0.923 \times 2392.8 = 2401\,\text{kJ}/
\text{kg}$, so the low-pressure stage delivers $h_5 - h_6 = 1177\,\text{kJ}
/\text{kg}$. The total turbine work is $546 + 1177 = 1723\,\text{kJ}/
\text{kg}$. The pump work, with $v_f = 0.00101$ and a $15\,\text{MPa}$
rise, is $w_p = 0.00101 \times (15000 - 10) = 15.1\,\text{kJ}/\text{kg}$,
so $h_2 = 191.8 + 15.1 = 207\,\text{kJ}/\text{kg}$. The reheat adds heat
twice: $q_h = (h_3 - h_2) + (h_5 - h_4) = (3448 - 207) + (3578 - 2902) =
3241 + 676 = 3917\,\text{kJ}/\text{kg}$. The thermal efficiency is
$\eta = (1723 - 15.1)/3917 = 0.436$, about $44\,\%$. Two results carry the
lesson. The exit quality $0.923$ comfortably clears the wet-steam erosion
limit, where the single-expansion cycle of the second worked example sat
at $0.82$; that protection of the last turbine stages is the first reason
to reheat. And the efficiency rose past the simple cycle's $37\,\%$
because the second heat addition lifts the average temperature at which
heat enters, the second reason. The reheat pressure is itself an
optimisation: too high and the second expansion is short, too low and the
reheat adds heat at a temperature barely above the condenser, so plants
place it near a quarter of the boiler pressure, as here.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-reheat-rankine-ts.tex}

\paragraph{Brayton with real component efficiencies.} Return to the
$r_p = 12$, $T_3 = 1400\,\si{\kelvin}$ Brayton case, but now let the
compressor and turbine fall short of isentropic, $\eta_c = 0.84$ and
$\eta_t = 0.88$, the values a working gas turbine reaches. The isentropic
endpoints are unchanged: $T_{2s} = 610\,\si{\kelvin}$ and $T_{4s} = 688\,
\si{\kelvin}$. The real compressor needs more work, so its exit is hotter,
$T_2 = T_1 + (T_{2s} - T_1)/\eta_c = 300 + 310/0.84 = 669\,\si{\kelvin}$,
and the real turbine delivers less, so its exit is hotter too, $T_4 = T_3
- \eta_t(T_3 - T_{4s}) = 1400 - 0.88 \times 712 = 773\,\si{\kelvin}$. The
turbine work falls to $c_p(T_3 - T_4) = 1.005 \times 627 = 630\,\text{kJ}
/\text{kg}$ and the compressor work rises to $c_p(T_2 - T_1) = 1.005
\times 369 = 371\,\text{kJ}/\text{kg}$, leaving a net work of only
$259\,\text{kJ}/\text{kg}$, down from the ideal $404$. The heat input
drops a little because the compressor delivers air already hotter,
$q_h = c_p(T_3 - T_2) = 1.005 \times 731 = 735\,\text{kJ}/\text{kg}$, so
$\eta = 259/735 = 0.352$. Two component efficiencies near $0.85$ pulled
the cycle efficiency from $51\,\%$ to $35\,\%$ and the net work down by a
third. The mechanism is the back-work ratio: with nearly half the turbine
output already committed to the compressor, a small fractional loss in
each machine bites twice, once as turbine output forgone and once as
compressor input demanded. This sensitivity is why gas-turbine makers
spend so heavily on compressor and turbine aerodynamics, and why the
back-work ratio, not the pressure ratio alone, is the number that governs
how forgiving the cycle is.

\paragraph{Two-stage refrigeration with intercooling.} A single-stage
vapour-compression cycle pulling a low evaporator temperature against a
warm condenser pays twice: the compressor works across a large pressure
ratio at low efficiency, and its discharge runs hot enough to stress the
oil and the valves. Splitting the compression into two stages with a
flash intercooler between them addresses both. Take R-134a with an
evaporator at $-30\,\degC$ and a condenser at $40\,\degC$, an intermediate
pressure chosen near the geometric mean of the evaporator and condenser
pressures (the rule that equalises the two stages' pressure ratios). The
evaporator pressure is $0.085\,\text{MPa}$ and the condenser pressure is
$1.017\,\text{MPa}$, so the geometric-mean intermediate is $\sqrt{0.085
\times 1.017} = 0.294\,\text{MPa}$, a saturation temperature near
$0\,\degC$. The low-stage compressor lifts vapour from $-30\,\degC$ to
$0\,\degC$; the flash chamber at $0\,\degC$ separates the liquid from the
flash vapour, the liquid feeds the evaporator through a second throttle,
and the flash vapour mixes with the low-stage discharge to de-superheat
it before the high-stage compressor lifts the combined flow to the
condenser. The mixing drops the high-stage inlet temperature, so the
high-stage discharge is cooler than a single stage's would be, and each
stage works across a pressure ratio of about $3.5$ rather than the single
stage's $12$. Carrying the enthalpies through the two-stage flow gives a
COP near $2.4$ against the single stage's $1.9$ for the same duty, a gain
of roughly a quarter, with the larger benefit the cooler discharge that
lets the machine run the low evaporator temperature at all. The
arithmetic rewards a careful mass balance at the flash chamber, since the
two stages carry different mass flows; that balance is the content of one
of the chapter's exercises.

\paragraph{The Stirling cycle and its regenerator.} The Stirling cycle reaches
the Carnot efficiency on paper through a device the Carnot cycle does not have,
an internal regenerator, and following the heat flows shows where the
efficiency is won and where a real machine loses it. The ideal cycle runs
between $T_h$ and $T_c$ through two isotherms and two constant-volume legs
(figure \ref{fig:vol04ch03:stirling-pv-ts}). Take a fixed mass of air,
$T_h = 900\,\si{\kelvin}$, $T_c = 300\,\si{\kelvin}$, a volume ratio
$V_{\max}/V_{\min} = 3$ across the isothermal legs, and the gas constant
$R = 0.287\,\text{kJ}/\text{kg}\cdot\text{K}$. The isothermal expansion at
$T_h$ absorbs $q_{12} = R T_h \ln(V_{\max}/V_{\min}) = 0.287 \times 900 \times
\ln 3 = 284\,\text{kJ}/\text{kg}$ from the hot reservoir; the isothermal
compression at $T_c$ rejects $q_{34} = R T_c \ln(V_{\max}/V_{\min}) = 0.287
\times 300 \times \ln 3 = 94.6\,\text{kJ}/\text{kg}$ to the cold reservoir.
The two constant-volume legs each move $c_v(T_h - T_c) = 0.718 \times 600 =
431\,\text{kJ}/\text{kg}$, one shedding it as the gas cools from $T_h$ to $T_c$,
the other taking it back as the gas heats from $T_c$ to $T_h$. With a perfect
regenerator those two transfers cancel internally, so the only heat crossing
the cycle boundary is the $284$ in and the $94.6$ out, and the efficiency is
$1 - 94.6/284 = 1 - T_c/T_h = 0.667$, the Carnot value. The lesson sits in the
size of the regenerated quantity: each constant-volume leg carries
$431\,\text{kJ}/\text{kg}$, larger than the net work of $284 - 94.6 = 189\,
\text{kJ}/\text{kg}$, so a regenerator that returns only $90\,\%$ of it leaks
$0.10 \times 431 = 43\,\text{kJ}/\text{kg}$ that the hot reservoir must then
resupply, dropping the efficiency from $66.7\,\%$ toward $189/(284 + 43) =
58\,\%$. The regenerator's effectiveness, not the isothermal legs, is the
number that governs a real Stirling engine, which is why the dead volume and
the thermal mass of the regenerator matrix dominate its design.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-stirling-pv-ts.tex}

\paragraph{Absorption refrigeration driven by heat.} An absorption chiller
replaces the compressor's mechanical work with a heat input, so its performance
is read against a different bound. The lithium-bromide and water machine common
in building cooling takes heat $q_{\text{gen}}$ into a generator at a driving
temperature $T_{\text{gen}}$, produces cooling $q_{\text{evap}}$ at the
evaporator temperature $T_{\text{evap}}$, and rejects heat at an intermediate
ambient $T_0$. The thermal coefficient of performance is
$\text{COP}_{\text{abs}} = q_{\text{evap}}/q_{\text{gen}}$, and the reversible
bound is the product of a Carnot heat-engine efficiency (extracting work from
the driving heat) and a Carnot refrigerator COP (using that work to cool),
\[
\text{COP}_{\text{abs,rev}} = \left(1 - \frac{T_0}{T_{\text{gen}}}\right)
\frac{T_{\text{evap}}}{T_0 - T_{\text{evap}}}.
\]
Take a single-effect machine driven by $90\,\degC$ hot water, $T_{\text{gen}} =
363\,\si{\kelvin}$, cooling to $T_{\text{evap}} = 280\,\si{\kelvin}$ ($7\,\degC$
chilled water) against $T_0 = 308\,\si{\kelvin}$ ($35\,\degC$ cooling tower).
The reversible bound is $(1 - 308/363)(280/28) = 0.152 \times 10.0 = 1.52$. A
real single-effect machine reaches a thermal COP near $0.7$, about $46\,\%$ of
the reversible bound, and a double-effect machine driven by steam reaches
$1.1$ to $1.4$. The number looks poor beside a vapour-compression COP of $4$,
but the comparison is unfair: the absorption machine consumes heat that would
otherwise be discarded, while the vapour-compression machine consumes
electricity worth several times its energy in fuel. Where waste heat or solar
heat is the input, the absorption machine's primary-energy performance can
exceed the electric machine's, which is the case section 3.6 makes for
trigeneration.

\paragraph{Regenerative gas turbine with intercooling and reheat.} The
intercooled, reheated, regenerative Brayton cycle approaches Carnot in the
limit of many stages, and one worked pass shows how far two stages of each
carries the simple cycle. Start from the $r_p = 12$, $T_1 = 300\,\si{\kelvin}$,
$T_3 = 1400\,\si{\kelvin}$ Brayton case, but split the compression into two
stages of pressure ratio $\sqrt{12} = 3.46$ each with intercooling back to
$300\,\si{\kelvin}$ between them, and split the expansion into two stages of
ratio $3.46$ each with reheat back to $1400\,\si{\kelvin}$ between them, then
add a regenerator. Each compressor stage now raises the temperature by
$T_1(3.46^{0.286} - 1) = 300 \times 0.425 = 127\,\si{\kelvin}$, so the total
compressor work is $2 \times c_p \times 127 = 2 \times 1.005 \times 127 =
256\,\text{kJ}/\text{kg}$, against the single stage's $312$. Each turbine stage
drops the temperature by $T_3(1 - 3.46^{-0.286}) = 1400 \times 0.298 =
417\,\si{\kelvin}$, so the total turbine work is $2 \times 1.005 \times 417 =
838\,\text{kJ}/\text{kg}$, against the single stage's $715$. The net work rises
to $838 - 256 = 582\,\text{kJ}/\text{kg}$, from the simple cycle's $403$, a gain
of $44\,\%$ in specific work from intercooling and reheat alone. Intercooling
and reheat by themselves lower the efficiency, because the intercooler rejects
heat and the reheat adds it at modest temperature, so the regenerator is
required to recover it: the turbine exhaust now leaves at a high temperature
(the last expansion stage starts hot), and a regenerator of effectiveness
$0.80$ preheats the combustor inlet with that exhaust, cutting the external
heat the combustor must supply. Carrying the regenerated heat balance through
lifts the cycle efficiency to roughly $0.55$, above the simple cycle's $0.51$
and with far more work per kilogram. The cost is three heat exchangers
(intercooler, reheat combustor, regenerator) and the pressure drops they add,
which is why aero-engines, where weight rules, decline the whole package and
stationary plants, where fuel cost rules, adopt it selectively.

\paragraph{Supercritical Rankine, where the dome disappears.} Modern coal
plants run above water's critical pressure, where the distinction between liquid
and vapour vanishes and the working fluid passes smoothly from compressed
liquid to superheated vapour with no two-phase plateau. Take a boiler at
$25\,\text{MPa}$, above the $22.1\,\text{MPa}$ critical pressure, superheated to
$600\,\degC$, against a $6\,\text{kPa}$ condenser, with an isentropic turbine.
From the superheated tables the turbine inlet has $h_3 = 3491\,\text{kJ}/\text{kg}$
and $s_3 = 6.36\,\text{kJ}/\text{kg}\cdot\text{K}$. Expanding isentropically to
$6\,\text{kPa}$, where $s_f = 0.521$ and $s_{fg} = 7.81$, gives an exit quality
$x_4 = (6.36 - 0.521)/7.81 = 0.748$ and $h_4 = 151.5 + 0.748 \times 2415 =
1958\,\text{kJ}/\text{kg}$, so the turbine delivers $h_3 - h_4 = 1533\,\text{kJ}/
\text{kg}$. The pump work, with $v_f = 0.00101$ across a $25\,\text{MPa}$ rise,
is $w_p = 0.00101 \times (25000 - 6) = 25.2\,\text{kJ}/\text{kg}$, so $h_2 =
151.5 + 25.2 = 177\,\text{kJ}/\text{kg}$ and $q_h = h_3 - h_2 = 3314\,\text{kJ}/
\text{kg}$. The thermal efficiency is $\eta = (1533 - 25.2)/3314 = 0.455$, about
$46\,\%$, against the $37\,\%$ of the $5\,\text{MPa}$ subcritical case of the
exercises. Two effects pay for the gain. The higher pressure lifts the average
temperature of heat addition, the same lever every Rankine refinement pulls; and
above the critical pressure the heat is added with no long constant-temperature
boiling plateau, so the average temperature of heat addition climbs further than
pressure alone would give. The cost is twofold: the exit quality $0.748$ is well
into the wet region, so this single-expansion supercritical cycle would erode its
last turbine stages and a real plant reheats once or twice to lift the exit
quality, exactly as the reheat example showed; and the $25\,\text{MPa}$ boiler
demands thick high-alloy tubing whose creep life at $600\,\degC$ is the binding
material constraint, the subject chapter~6 of this volume develops.

\paragraph{Transcritical \texorpdfstring{CO$_2$}{CO2} heat pump for hot water.}
Carbon dioxide refrigerant works above its low critical temperature
($31\,\degC$), so its high-side heat rejection is not a condensation at constant
temperature but a gas cooling that slides down a range of temperatures, which
suits water heating, where the water itself warms across a range. Take an
evaporator at $0\,\degC$ ($3.49\,\text{MPa}$) and a gas cooler running to a high
side of $9\,\text{MPa}$, above the $7.38\,\text{MPa}$ critical pressure, cooling
the high-pressure gas to $35\,\degC$ at exit. From CO$_2$ property data, the
evaporator-exit saturated vapour has $h_1 = 437\,\text{kJ}/\text{kg}$ and the
isentropic compression to $9\,\text{MPa}$ reaches $h_2 \approx 482\,\text{kJ}/
\text{kg}$, so the compressor work is $w_c = 45\,\text{kJ}/\text{kg}$. The gas
cooler takes the $9\,\text{MPa}$ stream from $h_2$ down to $h_3 = 264\,\text{kJ}/
\text{kg}$ at the $35\,\degC$ exit, rejecting $q_h = h_2 - h_3 = 218\,\text{kJ}/
\text{kg}$ across a glide from over $100\,\degC$ down to $35\,\degC$. The heating
COP is $q_h/w_c = 218/45 = 4.8$. The number rests on a low gas-cooler exit
temperature, which is why the transcritical CO$_2$ cycle excels at heating cold
inlet water to a high delivery temperature: the cold water pulls the gas-cooler
exit down, and a low exit raises both the rejected heat and the COP. Feeding the
gas cooler with already-warm return water raises the exit temperature and
collapses the COP, the failure mode that distinguishes a well-plumbed CO$_2$
water heater from a poorly plumbed one. The cycle pays for its low global-warming
working fluid with operating pressures several times those of a conventional
machine, which sets the wall thickness of every component.

\paragraph{Diesel cycle with realistic peak-pressure limit.} The Diesel
efficiency formula rewards a high compression ratio, but a real engine caps the
peak cylinder pressure to keep the structure and bearings within their limits,
and that cap, not the compression ratio alone, often sets the achievable cycle.
Take a compression ratio $r = 18$, inlet $T_1 = 310\,\si{\kelvin}$ and $p_1 =
100\,\text{kPa}$, $\gamma = 1.4$. The compression gives $T_2 = 310 \times
18^{0.4} = 985\,\si{\kelvin}$ and $p_2 = 100 \times 18^{1.4} = 5720\,\text{kPa}$.
Constant-pressure combustion holds the pressure at $5.72\,\text{MPa}$ through the
heat addition, so the peak pressure is the compression pressure, $5.72\,\text{MPa}$,
comfortably within a $16\,\text{MPa}$ structural limit. Suppose the injected fuel
raises the temperature to $T_3 = 2000\,\si{\kelvin}$; the cut-off ratio is then
$r_c = T_3/T_2 = 2000/985 = 2.03$. The efficiency is $\eta_D = 1 - (1/18^{0.4})
(2.03^{1.4} - 1)/(1.4 \times 1.03) = 1 - 0.314 \times (2.70 - 1)/1.44 = 1 -
0.314 \times 1.18 = 0.629$, about $63\,\%$ ideal. The lesson is the comparison
the structural cap forces: a spark-ignition engine adds its heat at constant
volume, so its peak pressure climbs with the heat added and an aggressive Otto
design slams against the same structural ceiling at a far lower compression
ratio, while the Diesel's constant-pressure burn lets it raise the compression
ratio and the efficiency without raising the peak pressure beyond the
compression value. The constant-pressure combustion that costs the Diesel a few
points of cycle efficiency at equal compression ratio is the same feature that
lets it run the high compression ratio safely, the trade that the section 3.4
text described and this peak-pressure accounting makes quantitative.

\paragraph{Throttling loss versus an expander, by exergy.} The vapour-compression
example named the throttle as the cycle's main irreversibility; the exergy
accounting puts a number on what an expander would recover. Return to the R-134a
cycle with the condenser at $40\,\degC$ ($313\,\si{\kelvin}$) and the evaporator
at $-10\,\degC$ ($263\,\si{\kelvin}$), saturated liquid leaving the condenser at
$h_3 = 256.4\,\text{kJ}/\text{kg}$, $s_3 = 1.1909\,\text{kJ}/\text{kg}\cdot\text{K}$.
The throttle holds enthalpy, $h_4 = h_3$, but it generates entropy: the
evaporator-pressure state at $h_4$ has $s_4 = 1.2110\,\text{kJ}/\text{kg}\cdot
\text{K}$, so the entropy generated is $s_{\text{gen}} = s_4 - s_3 = 0.0201\,
\text{kJ}/\text{kg}\cdot\text{K}$. With a dead state at $T_0 = 313\,\si{\kelvin}$,
the lost work is $w_{\text{lost}} = T_0 s_{\text{gen}} = 313 \times 0.0201 =
6.3\,\text{kJ}/\text{kg}$, the Gouy-Stodola figure of the chapter's first mastery
box applied to one component. Set that against the compressor work of $33.2\,
\text{kJ}/\text{kg}$: the throttle alone forfeits work potential equal to about
$19\,\%$ of the compressor input. An isentropic expander replacing the throttle
would recover that potential and, by ending the expansion at lower enthalpy,
would also raise the cooling effect $q_c$, lifting the cycle COP from $4.1$
toward roughly $4.6$. The reason production refrigerators keep the cheap throttle
despite this is that a two-phase expander recovering a few kilojoules per kilogram
is mechanically intricate and rarely pays its capital back at domestic scale;
large industrial plants and some CO$_2$ systems, where the throttling loss is a
larger share, do fit expanders, which is the design line the exergy number draws.

\paragraph{Brayton specific-work optimum, worked at two ratios.} The
specific-work formula $w_{\text{net}}/(c_p T_1) = \tau(1 - 1/x) - (x - 1)$,
with $\tau = T_3/T_1$ and $x = r_p^{(\gamma-1)/\gamma}$, locates the pressure
ratio that makes a gas turbine smallest for a given power, and a numerical
pass fixes how sharply the work falls away from that optimum. Take $T_1 =
300\,\si{\kelvin}$, $T_3 = 1500\,\si{\kelvin}$, so $\tau = 5$, and $\gamma =
1.4$. The optimum is $x^\star = \sqrt{5} = 2.236$, that is $r_p^\star =
2.236^{1/0.286} = 2.236^{3.5} = 16.7$, and there the normalised net work is
$5(1 - 1/2.236) - (2.236 - 1) = 5 \times 0.553 - 1.236 = 1.53$, so with $c_p
T_1 = 1.005 \times 300 = 302\,\text{kJ}/\text{kg}$ the peak specific work is
$1.53 \times 302 = 462\,\text{kJ}/\text{kg}$. Step the pressure ratio to
$r_p = 30$, where $x = 30^{0.286} = 2.62$: the normalised work falls to
$5(1 - 1/2.62) - 1.62 = 3.09 - 1.62 = 1.47$, only $4\,\%$ below the peak, yet
the thermal efficiency has risen from $1 - 1/2.236 = 0.553$ at the optimum to
$1 - 1/2.62 = 0.618$. The flatness of the specific-work curve near its peak
is the engineer's room to manoeuvre: moving from $r_p = 16.7$ to $r_p = 30$
buys six points of efficiency for a four-percent loss of specific work, which
is why stationary machines sit above the specific-work optimum. The companion
\texttt{brayton\_cycle.py} returns these temperatures and works for any
$(r_p, T_3)$ pair, and figure \ref{fig:vol04ch03:brayton-specific-work} plots
the family the two points sample.

\paragraph{Regenerator effectiveness and the heat it saves.} Return to the
$r_p = 12$, $T_3 = 1400\,\si{\kelvin}$, $T_1 = 300\,\si{\kelvin}$ Brayton case
of the earlier worked example, with isentropic components for clarity, and add
a regenerator of effectiveness $\varepsilon = 0.80$. The compressor exit is
$T_2 = 610\,\si{\kelvin}$ and the turbine exit is $T_4 = 688\,\si{\kelvin}$,
so the temperature gap the regenerator works across is only $T_4 - T_2 =
78\,\si{\kelvin}$, narrow because $r_p = 12$ sits close to the crossover ratio
$\tau^{\gamma/[2(\gamma-1)]} = 4.67^{1.75} = 14.8$. The regenerator raises the
combustor inlet to $T_x = T_2 + \varepsilon(T_4 - T_2) = 610 + 0.80 \times 78
= 672\,\si{\kelvin}$, so the combustor now supplies $q_h = c_p(T_3 - T_x) =
1.005 \times 728 = 732\,\text{kJ}/\text{kg}$ instead of the simple cycle's
$794\,\text{kJ}/\text{kg}$, a saving of $62\,\text{kJ}/\text{kg}$. The net
work is unchanged at $403\,\text{kJ}/\text{kg}$, so the efficiency rises to
$403/732 = 0.55$, from the simple cycle's $0.51$. The lift is modest because
the pressure ratio is near the crossover, where the regenerator has little gap
to feed on; the same regenerator on a $r_p = 6$ machine, where $T_4 - T_2$ is
several times wider, would lift the efficiency far more, the design region
figure \ref{fig:vol04ch03:regeneration-efficiency} marks as where regeneration
pays.

\paragraph{Optimum intercooling pressure, two stages.} A gas turbine
compresses air from $100\,\text{kPa}$ to $1600\,\text{kPa}$, an overall
pressure ratio of $16$, and the question is where to set the intermediate
pressure for two-stage compression with intercooling back to the
$300\,\si{\kelvin}$ inlet temperature. The geometric-mean rule gives
$p_i = \sqrt{100 \times 1600} = 400\,\text{kPa}$, so each stage carries the
ratio $\sqrt{16} = 4$. With perfect intercooling each stage starts at
$300\,\si{\kelvin}$ and raises the temperature by $T_1(4^{0.286} - 1) = 300
\times (1.486 - 1) = 146\,\si{\kelvin}$, so the two-stage compressor work is
$2 \times c_p \times 146 = 2 \times 1.005 \times 146 = 293\,\text{kJ}/
\text{kg}$. Single-stage compression to the same $16:1$ ratio would reach
$T_2 = 300 \times 16^{0.286} = 300 \times 2.21 = 663\,\si{\kelvin}$, costing
$c_p(663 - 300) = 1.005 \times 363 = 365\,\text{kJ}/\text{kg}$, so the
intercooler saves $72\,\text{kJ}/\text{kg}$ of compressor work, a fifth of
the single-stage figure. To confirm the geometric mean is the optimum and not
merely a convenient choice, try $p_i = 300\,\text{kPa}$: the first stage
ratio is $3$ and the second $5.33$, raising temperatures by $300(3^{0.286} -
1) = 111\,\si{\kelvin}$ and $300(5.33^{0.286} - 1) = 184\,\si{\kelvin}$, for a
total work $1.005 \times 295 = 296\,\text{kJ}/\text{kg}$, marginally worse
than the balanced $293$. The curve is flat near its minimum, as every
geometric-mean optimum is, but the balanced split wins, and it also equalises
the discharge temperatures, which eases the mechanical design of the two
stages.

\paragraph{Cascade refrigeration for a low-temperature load.} A single
refrigerant cannot span a very deep lift: the fluid that boils at a low
enough temperature for the cold load condenses at an impractically high
pressure at the warm rejection temperature. A \engterm{cascade} cycle stacks
two loops with different refrigerants, the cold loop's condenser cooled by the
warm loop's evaporator through a shared heat exchanger. Take a freezer load at
$-40\,\degC$ rejecting to a $35\,\degC$ ambient. A single R-134a loop across
that span faces a pressure ratio near $\exp[(h_{fg}/R)(1/233 - 1/308)]
\approx 18$, deep into the region where compressor efficiency collapses.
Split it: a low-temperature loop using a fluid suited to $-40\,\degC$ rejects
its heat at, say, $-5\,\degC$ to the high-temperature loop's evaporator, and
the high-temperature R-134a loop carries that heat from $-10\,\degC$ up to the
$35\,\degC$ condenser. Each loop now works across a lift near $40\,\si{
\kelvin}$ and a pressure ratio near $4$, in the efficient range, with a
$5\,\si{\kelvin}$ pinch in the cascade heat exchanger as the price of the
split. Carrying representative enthalpies through both loops gives an overall
COP near $1.3$ for the $-40\,\degC$ duty, against the fraction of that a
strangled single loop would manage, and the deeper lesson is the
multiplicative one: the overall COP of the cascade is built from the two
loops' COPs through the heat they pass between them, so adding a third stage
for a cryogenic load extends the same logic the two-stage and cascade
arrangements share with the staged compression of the Brayton cycle.

\paragraph{Ericsson cycle, the other regenerated Carnot equal.} The Stirling
example reached the Carnot efficiency through congruent constant-volume legs; the
Ericsson cycle reaches it through congruent constant-pressure legs, and setting
the two side by side shows that the regenerator, not the shape of the connecting
process, is what wins the efficiency. The ideal Ericsson cycle runs two isotherms
at $T_h$ and $T_c$ closed by two isobars (figure
\ref{fig:vol04ch03:ericsson-ts}). Take air, $T_h = 900\,\si{\kelvin}$, $T_c =
300\,\si{\kelvin}$, an isothermal pressure ratio of $6$ across each isotherm, and
$R = 0.287\,\text{kJ}/\text{kg}\cdot\text{K}$. The isothermal expansion at $T_h$
absorbs $q_{h} = R T_h \ln 6 = 0.287 \times 900 \times 1.792 = 463\,\text{kJ}/
\text{kg}$, and the isothermal compression at $T_c$ rejects $q_{c} = R T_c \ln 6
= 0.287 \times 300 \times 1.792 = 154\,\text{kJ}/\text{kg}$. The two isobaric
legs each move $c_p(T_h - T_c) = 1.005 \times 600 = 603\,\text{kJ}/\text{kg}$,
one shedding it as the gas cools at high pressure, the other taking it back as the
gas heats at low pressure, and because the isobars are congruent those two
transfers are equal and a counter-flow regenerator carries the $603$ internally.
With the regeneration perfect, the only heat crossing the boundary is the $463$ in
and the $154$ out, and the efficiency is $1 - 154/463 = 0.667$, once again the
Carnot value $1 - T_c/T_h$. The regenerated quantity $603\,\text{kJ}/\text{kg}$
again exceeds the net work $463 - 154 = 309\,\text{kJ}/\text{kg}$, so a
regenerator returning only $90\,\%$ of it forces the hot reservoir to resupply
$0.10 \times 603 = 60\,\text{kJ}/\text{kg}$ and drops the efficiency toward
$309/(463 + 60) = 59\,\%$. The Stirling and Ericsson cycles are therefore the
same idea in two geometries: isotherms plus a perfect regenerator reach Carnot,
and the regenerator effectiveness is the number a real machine lives or dies by,
whichever legs connect the isotherms.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-ericsson-ts.tex}

\paragraph{Turbofan, where thermal efficiency meets propulsive efficiency.} A
jet engine converts fuel heat to a thrust that pushes an aircraft, and its useful
output is not shaft work but the rate at which thrust does work on the airframe,
so its overall efficiency is a product of two terms the earlier Brayton examples
never separated. The core is a Brayton cycle of thermal efficiency
$\eta_{\text{th}}$, delivering kinetic energy to a jet; the propulsion stage turns
that jet's kinetic energy into thrust work at a \engterm{propulsive efficiency}
\[
\eta_p = \frac{2}{1 + v_j/v_0},
\]
where $v_j$ is the exhaust velocity and $v_0$ the flight velocity (figure
\ref{fig:vol04ch03:propulsive-efficiency}). The overall efficiency is
$\eta_o = \eta_{\text{th}}\,\eta_p$. Take a subsonic transport cruising at
$v_0 = 250\,\text{m}/\text{s}$. A bare turbojet throws its jet at
$v_j = 600\,\text{m}/\text{s}$, so $v_j/v_0 = 2.4$ and $\eta_p = 2/3.4 = 0.59$;
a high-bypass turbofan gives a large mass of air a small increment, exhausting at
$v_j = 320\,\text{m}/\text{s}$, so $v_j/v_0 = 1.28$ and $\eta_p = 2/2.28 = 0.88$.
With a core thermal efficiency near $0.45$ in both cases, the turbojet reaches an
overall $0.45 \times 0.59 = 0.27$ while the turbofan reaches $0.45 \times 0.88 =
0.40$, half again as efficient at the same cruise for the same core. The thrust
per unit mass flow follows the same accounting: thrust is $\dot m(v_j - v_0)$ and
the propulsive power is $\text{thrust} \times v_0$, so throwing a small mass fast
wastes energy in the wake, the kinetic energy $\tfrac12 \dot m(v_j - v_0)^2$ the
aircraft never recovers. The design reading is that at subsonic speed the way to
raise efficiency is to lower $v_j$ toward $v_0$ by moving more air more gently,
which is the whole logic of the bypass ratio's steady climb, and it explains why
turbofans supplanted turbojets on airliners while pure jets survive only where
supersonic flight forces $v_j$ high.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-propulsive-efficiency.tex}

\paragraph{Linde-Hampson liquefier, cold from a throttle alone.} A gas-cycle
refrigerator can reach cryogenic temperature with no low-temperature moving part
by exploiting the Joule-Thomson effect: a real gas throttled from high to low
pressure below its inversion temperature cools, and recycling the cold gas to
pre-cool the incoming stream drives the process down to liquefaction. Take a
Linde-Hampson air liquefier compressing to $p_2 = 20\,\text{MPa}$ and expanding
through a throttle to $p_1 = 0.1\,\text{MPa}$, with the make-up gas entering at
$T_1 = 300\,\si{\kelvin}$. An energy balance on the whole cold box, drawn around
the counter-flow heat exchanger, the throttle, and the liquid receiver, gives the
liquid yield per unit of gas compressed. With $h_1$ the incoming enthalpy at
$300\,\si{\kelvin}$ and $p_1$, $h_2$ the enthalpy at $300\,\si{\kelvin}$ and
$p_2$, and $h_f$ the saturated-liquid enthalpy at $p_1$, the fraction liquefied is
\[
y = \frac{h_1 - h_2}{h_1 - h_f}.
\]
For air the isothermal throttling enthalpy drop $h_1 - h_2$ across the
$0.1$-to-$20\,\text{MPa}$ span at $300\,\si{\kelvin}$ is about $40\,\text{kJ}/
\text{kg}$, while $h_1 - h_f$, the enthalpy that must be removed to take
atmospheric-pressure air from $300\,\si{\kelvin}$ to saturated liquid at
$79\,\si{\kelvin}$, is about $430\,\text{kJ}/\text{kg}$, so $y = 40/430 = 0.093$:
under a tenth of the compressed air leaves as liquid, and the rest returns
through the exchanger to cool the next batch. The small yield is the signature of
the simple Linde cycle, and it is why industrial liquefiers add a work-extracting
expander (the Claude cycle) that cools the gas by making it do work as well as by
throttling, lifting the yield severalfold. The lesson the yield formula carries
is that the entire process rests on the sign of the Joule-Thomson coefficient:
above the inversion temperature the throttle warms the gas and $y$ goes negative,
which is why hydrogen and helium liquefiers must pre-cool below their low
inversion temperatures before the same throttle will cool at all.

\paragraph{Rankine with one open feedwater heater.} Regenerative feedwater
heating raises the average temperature of heat addition by bleeding steam from the
turbine to warm the feedwater, and a single open heater carried through with
numbers shows the mechanism and the mass-fraction bookkeeping it turns on. Take
the $5\,\text{MPa}/450\,\degC$ boiler and $10\,\text{kPa}$ condenser of the
chapter's first Rankine exercise, but bleed a fraction $y$ of the steam at an
intermediate $0.5\,\text{MPa}$ into an open feedwater heater, where it mixes
directly with the condensate and leaves as saturated liquid at $0.5\,\text{MPa}$.
From the tables the turbine inlet is $h_3 = 3317\,\text{kJ}/\text{kg}$,
$s_3 = 6.82\,\text{kJ}/\text{kg}\cdot\text{K}$; expanding to $0.5\,\text{MPa}$
reaches $h_6 \approx 2758\,\text{kJ}/\text{kg}$, and the extraction bleeds steam
there. The heater energy balance, per unit boiler flow, sets the bled fraction:
$y\,h_6 + (1 - y)\,h_2 = h_7$, with $h_2$ the condensate pumped to
$0.5\,\text{MPa}$ ($h_2 \approx 192\,\text{kJ}/\text{kg}$) and $h_7$ the
saturated liquid leaving the heater ($h_7 = 640\,\text{kJ}/\text{kg}$ at
$0.5\,\text{MPa}$). Solving, $y = (h_7 - h_2)/(h_6 - h_2) = (640 - 192)/(2758 -
192) = 448/2566 = 0.175$: about a sixth of the steam is diverted to the heater.
That diverted steam gives up its heat to the feedwater instead of expanding to the
cold condenser, so the boiler now takes water in at $640\,\text{kJ}/\text{kg}$
rather than $197\,\text{kJ}/\text{kg}$, cutting the high-temperature heat it must
supply. Carrying the reduced turbine work (the bled sixth expands only partway)
against the reduced heat input lifts the cycle efficiency from the plain cycle's
$37\,\%$ to about $39\,\%$. The gain is modest for one heater and grows with the
number of stages, which is why large plants use six to ten; the bookkeeping
lesson is that every regenerative stage adds a mass-balance unknown, the bled
fraction, fixed by an energy balance on its heater.

\paragraph{Thermoelectric cooler and the figure of merit.} A Peltier module moves
heat with no working fluid and no moving part by driving current through a
junction of two semiconductors, and its performance is governed by a single
dimensionless material group rather than by a cycle geometry. The maximum cooling
COP of an optimised module between a cold side $T_c$ and a hot side $T_h$ is the
Carnot COP scaled by a figure-of-merit factor,
\[
\text{COP}_{\max} = \frac{T_c}{T_h - T_c}\,
\frac{\sqrt{1 + Z\bar T} - T_h/T_c}{\sqrt{1 + Z\bar T} + 1},
\]
where $\bar T = (T_h + T_c)/2$ and $Z\bar T$ is the dimensionless figure of merit
(figure \ref{fig:vol04ch03:thermoelectric-cop}). Take a module holding a cold
side at $0\,\degC$ ($273\,\si{\kelvin}$) against a hot side at $27\,\degC$
($300\,\si{\kelvin}$), a lift of $27\,\si{\kelvin}$, so the Carnot COP is
$273/27 = 10.1$. A commercial bismuth-telluride module reaches $Z\bar T \approx
1$, so $\sqrt{1 + Z\bar T} = \sqrt 2 = 1.414$, and with $T_h/T_c = 300/273 =
1.099$ the factor is $(1.414 - 1.099)/(1.414 + 1) = 0.315/2.414 = 0.130$, giving
$\text{COP}_{\max} = 10.1 \times 0.130 = 1.3$. The module reaches about an eighth
of the Carnot bound, an order of magnitude worse than the vapour-compression
cycle's $4.1$ at a comparable duty. Reaching a Carnot fraction of one-half would
take $Z\bar T \approx 4$, and no bulk material has crossed $Z\bar T = 3$. The
figure of merit collects the material's Seebeck coefficient, electrical
conductivity, and thermal conductivity into one number, and the reason it is hard
to raise is that the three properties fight one another: raising electrical
conductivity tends to raise thermal conductivity, which the denominator of $Z$
penalises. Thermoelectric cooling therefore survives not on efficiency but where
its silence, its absence of refrigerant, and its ability to shrink to a
millimetre scale outrank the COP, in instrument cooling, portable coolers, and
laser-diode temperature control.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-thermoelectric-cop.tex}

\paragraph{Steam-jet ejector refrigeration, cold from a pressure nozzle.} An
ejector chiller replaces the compressor with a steam nozzle, so like the
absorption machine it is driven by heat rather than shaft work, and its
performance is read against the same product-of-Carnot-factors bound. High-pressure
motive steam expands through a converging-diverging nozzle to supersonic speed,
and the low pressure it creates entrains vapour from an evaporator, pulling the
cooling load; the mixed stream then recompresses through a diffuser to the
condenser pressure. Take motive steam generated at $130\,\degC$
($T_{\text{gen}} = 403\,\si{\kelvin}$), an evaporator at $7\,\degC$
($T_{\text{evap}} = 280\,\si{\kelvin}$) chilling water, and heat rejection to a
condenser at $35\,\degC$ ($T_0 = 308\,\si{\kelvin}$). The reversible bound is the
product of a Carnot heat-engine efficiency on the motive steam and a Carnot
refrigerator COP,
\[
\text{COP}_{\text{ej,rev}} = \left(1 - \frac{T_0}{T_{\text{gen}}}\right)
\frac{T_{\text{evap}}}{T_0 - T_{\text{evap}}} =
\left(1 - \frac{308}{403}\right)\frac{280}{28} = 0.236 \times 10.0 = 2.4.
\]
A real steam-jet chiller reaches a thermal COP near $0.3$ to $0.5$, only a fifth
of the reversible bound, because the nozzle and diffuser destroy a large share of
the motive steam's exergy in the supersonic mixing of two streams at very
different velocities. That mixing loss is the ejector's defining weakness and the
reason it lost ground to absorption chillers, which reach thermal COPs of $0.7$
and above for the same duty. The ejector survives where its virtues, no moving
parts, tolerance of a two-phase working stream, and the ability to run on
low-grade steam or even the vapour a process already produces, matter more than
the COP, above all in large water-chilling plants where the working fluid is water
itself and the driving steam is a byproduct.

\paragraph{Dual (Sabath\'e) cycle, between Otto and Diesel.} A real
compression-ignition engine burns neither at constant volume like the ideal
Otto nor at constant pressure like the ideal Diesel; its charge burns partly at
constant volume as the premixed fraction ignites and partly at constant pressure
as the rest injects, and the dual cycle captures that with two heat-addition
legs (figure \ref{fig:vol04ch03:dual-cycle-pv}). Take a compression ratio
$r = 18$, inlet $T_1 = 300\,\si{\kelvin}$, $\gamma = 1.4$, and split the heat so
that a constant-volume leg raises the pressure by a factor $r_p = 1.5$ before a
constant-pressure leg carries a cut-off ratio $r_c = 1.6$. Compression gives
$T_2 = 300 \times 18^{0.4} = 953\,\si{\kelvin}$. The constant-volume leg raises
the temperature in proportion to the pressure, $T_x = r_p T_2 = 1.5 \times 953 =
1430\,\si{\kelvin}$; the constant-pressure leg raises it in proportion to the
volume, $T_3 = r_c T_x = 1.6 \times 1430 = 2288\,\si{\kelvin}$. The air-standard
dual efficiency has the closed form
\[
\eta_{\text{dual}} = 1 - \frac{1}{r^{\gamma-1}}\,
\frac{r_p r_c^\gamma - 1}{(r_p - 1) + \gamma\, r_p (r_c - 1)},
\]
which returns the Otto expression at $r_c = 1$ (all heat at constant volume) and
the Diesel expression at $r_p = 1$ (all heat at constant pressure). Substituting,
the bracket numerator is $1.5 \times 1.6^{1.4} - 1 = 1.5 \times 1.90 - 1 = 1.85$
and the denominator is $(1.5 - 1) + 1.4 \times 1.5 \times (1.6 - 1) = 0.5 + 1.26
= 1.76$, so with $18^{0.4} = 3.18$,
$\eta_{\text{dual}} = 1 - (1/3.18)(1.85/1.76) = 1 - 0.314 \times 1.05 = 0.670$,
about $67\,\%$. The value sits between the pure Diesel at this compression ratio
and cut-off (near $63\,\%$) and the pure Otto (near $69\,\%$), and shifting the
split toward more constant-volume burn, a larger $r_p$, walks the efficiency
toward the Otto value while raising the peak pressure the constant-volume leg
produces. The dual cycle is the honest air-standard model of a modern diesel,
and its two parameters are the knobs an injection strategy actually turns.

\paragraph{Zeotropic mixture and the temperature glide it buys.} A pure working
fluid boils at one temperature, so in a boiler heated by a stream that is itself
cooling the two curves cross at a single pinch and open a wide gap everywhere
else, across which heat is transferred irreversibly. A zeotropic mixture, two
fluids of different volatility, boils across a range of temperatures, a
\engterm{glide}, and matching that glide to the source recovers the lost work
(figure \ref{fig:vol04ch03:glide-matching}). Take the landfill-gas waste stream
of the organic-Rankine case study, cooling from $150\,\degC$ to a stack at
$90\,\degC$ across the boiler, and compare a pure fluid boiling flat at
$120\,\degC$ against a mixture gliding from $105\,\degC$ to $140\,\degC$. The
pure fluid's mean temperature of heat addition, weighted by the heat taken in,
sits near $120\,\degC$ ($393\,\si{\kelvin}$); the gliding mixture's mean sits
near $\tfrac12(105 + 140) = 122.5\,\degC$ ($395\,\si{\kelvin}$), only two kelvin
higher on the working-fluid side, but the source side tells the real story. The
average temperature gap between source and working fluid falls from roughly
$25\,\si{\kelvin}$ for the pure fluid to roughly $12\,\si{\kelvin}$ for the
matched glide, and the entropy generated in the boiler scales with that gap. For
a boiler duty of $3630\,\text{kW}$ the entropy generation drops from about
$\dot Q(1/T_{\text{wf}} - 1/T_{\text{src}}) \approx 3630(1/393 - 1/418) =
0.55\,\text{kW}/\si{\kelvin}$ to about $3630(1/395 - 1/407) = 0.27\,\text{kW}/
\si{\kelvin}$, and the lost work $T_0 \dot S_{\text{gen}}$ with $T_0 =
308\,\si{\kelvin}$ falls from $170\,\text{kW}$ to $83\,\text{kW}$, roughly
$87\,\text{kW}$ of work potential recovered on a half-megawatt machine. The gain
is the Kalina cycle's whole premise, an ammonia-water mixture whose glide is
tuned to the source, and it is the same average-temperature lever every Rankine
refinement pulls, applied to the shape of the heat-addition curve rather than to
its endpoints. The cost is that the mixture composition drifts if the two
components leak at different rates, and that condensation now also glides, which
complicates the condenser as much as the glide simplifies the boiler.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-dual-cycle-pv.tex}

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-glide-matching.tex}

\paragraph{Supercritical CO\texorpdfstring{$_2$}{2} recompression Brayton cycle.}
Carbon dioxide held above its critical point makes a compact power cycle because
compressing it near the critical point, where the fluid is dense, costs little
work, so the back-work penalty that burdens a gas Brayton cycle nearly vanishes
(figure \ref{fig:vol04ch03:sco2-recompression-ts}). Take a simple recuperated
sCO$_2$ cycle with the compressor inlet at $32\,\degC$ and $8\,\text{MPa}$, just
above the $31\,\degC$ critical temperature, compressed to $25\,\text{MPa}$, and a
turbine inlet at $600\,\degC$. Near the critical point the CO$_2$ density is about
$400\,\text{kg}/\text{m}^3$, an order of magnitude above an ideal gas at the same
pressure, so the compression takes only $w_c \approx 22\,\text{kJ}/\text{kg}$
against a turbine output near $w_t \approx 150\,\text{kJ}/\text{kg}$, a back-work
ratio of $0.15$ where an air Brayton cycle at similar conditions runs $0.4$ to
$0.5$. The simple recuperated cycle reaches a thermal efficiency near $43\,\%$;
the recompression variant, which splits the flow so a recompressor bypasses the
precooler and lets the recuperators absorb the mismatch in heat capacity between
the high-pressure and low-pressure streams, lifts it toward $48\,\%$ to $50\,\%$
at these conditions. The heat capacity mismatch is the crux the recompression
solves: near the critical point the high-pressure stream has a far larger
specific heat than the low-pressure stream, so a single recuperator would pinch
internally and stall the heat recovery, and diverting a fraction of the flow
around the precooler rebalances the two duties. The cycle's appeal is its
size, a sCO$_2$ turbine for a given power is a fraction of a steam turbine's
volume, which is why it is pursued for nuclear, concentrated-solar, and
ship-board power where the steam cycle's bulk and water demand are liabilities;
the cost is that $25\,\text{MPa}$ CO$_2$ at $600\,\degC$ is corrosive and demands
the same high-alloy discipline the supercritical steam plant needs.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-sco2-recompression-ts.tex}

\paragraph{Trilateral flash cycle, where boiling is skipped.} For a waste stream
that cools steeply, even a gliding boiler wastes work, because the fluid still
spends its heat intake climbing to a boiling temperature the source races past.
The trilateral flash cycle answers with a working fluid heated only as a liquid,
to saturation and no further, then flashed across a two-phase expander, so the
heat-addition curve is a single sensible-heating line that can hug a steeply
cooling source almost perfectly. Take a $120\,\degC$ pressurised-water waste
stream cooling to $60\,\degC$. A liquid working fluid heated from $60\,\degC$ to
$115\,\degC$ tracks that source within a near-constant $5\,\si{\kelvin}$ gap the
whole way, so the boiler entropy generation is a fraction of an isothermal
boiler's, and the second-law efficiency against the source can exceed a
conventional organic Rankine cycle's by ten to twenty percent. The catch is the
expander: flashing a saturated liquid produces a two-phase mixture that begins at
zero quality and ends perhaps thirty percent vapour, and expanding that mixture
with any efficiency demands a two-phase machine, a screw or scroll expander
tolerant of liquid, not a clean vapour turbine. The trilateral cycle therefore
trades a superior heat-source match for an inferior expander, and it pays only
where the source is both low-grade and steeply cooling, the regime that defeats
the isothermal-boiling cycles and rewards the fluid that never boils in the
boiler at all.

\paragraph{Aircraft air-cycle machine, bootstrap configuration.} An aircraft
environmental-control pack runs a reversed Brayton cycle on engine bleed air, and
the bootstrap arrangement uses the cold expander's own shaft work to drive a
second compressor that raises the pressure before expansion, deepening the cold.
Take bleed air entering the pack at $200\,\degC$ ($473\,\si{\kelvin}$) and
$250\,\text{kPa}$. A primary heat exchanger against ram air cools it to
$70\,\degC$ ($343\,\si{\kelvin}$); the bootstrap compressor then raises it to
$400\,\text{kPa}$, heating it to $T = 343 \times (400/250)^{0.286} =
343 \times 1.144 = 392\,\si{\kelvin}$; a secondary ram-air exchanger cools it
back to $70\,\degC$ ($343\,\si{\kelvin}$); and the cold-air turbine expands it to
the $85\,\text{kPa}$ cabin pressure, reaching $T = 343 \times (85/400)^{0.286} =
343 \times 0.647 = 222\,\si{\kelvin}$, about $-51\,\degC$, with an isentropic
turbine, or near $-35\,\degC$ once a turbine efficiency of $0.80$ is applied.
That sub-freezing air mixes with recirculated cabin air to hold the cabin near
$22\,\degC$, and the turbine's extracted work drives the bootstrap compressor, so
the pack needs no external shaft power beyond the bleed air the engine already
provides. The reversed Brayton COP here is low, of order one, but the cycle earns
its place on three counts the vapour-compression machine cannot match: the
working fluid is the cabin air itself, so a leak is harmless; the air arrives
already pressurised and breathable from the bleed system; and the whole pack is
light and free of any refrigerant charge to service, which on an aircraft
outranks the COP the ground-based machine would demand.

\paragraph{Regenerative Rankine with a closed feedwater heater.} The open
feedwater heater of the earlier worked example mixed the bled steam directly into
the feedwater and so needed a pump after every heater; a \engterm{closed}
feedwater heater keeps the bled steam on the other side of a tube wall, drains
its condensate to a lower-pressure heater or the condenser, and lets one
feed pump serve the whole train. Take the $5\,\text{MPa}/450\,\degC$ boiler and
$10\,\text{kPa}$ condenser again, bleeding steam at $0.5\,\text{MPa}$
($h_6 = 2758\,\text{kJ}/\text{kg}$) into a closed heater that warms the
high-pressure feedwater to the bled steam's saturation temperature,
$152\,\degC$, at which the feedwater enthalpy is about $h_{\text{fw,out}} =
645\,\text{kJ}/\text{kg}$, up from the $197\,\text{kJ}/\text{kg}$ the pump
delivered. The bled steam condenses to saturated liquid at $0.5\,\text{MPa}$,
$h_{\text{drain}} = 640\,\text{kJ}/\text{kg}$, and drains to the condenser. The
heater energy balance sets the bled fraction: $y(h_6 - h_{\text{drain}}) =
(1 - y)(h_{\text{fw,out}} - h_{\text{fw,in}})$, so $y(2758 - 640) =
(1 - y)(645 - 197)$, giving $2118\,y = 448(1 - y)$, $y = 448/2566 = 0.175$,
the same sixth of the flow the open heater diverted. The closed heater's
distinction is not the fraction but the plumbing: the drained condensate carries
its $640\,\text{kJ}/\text{kg}$ down to the condenser rather than up to the boiler,
a small thermodynamic penalty against the open heater's direct mixing, paid to
avoid a pump at every extraction pressure. Real plants use both kinds, closed
heaters for most stages because they need only drains, and one open heater
doubling as a deaerator to strip dissolved gases from the feedwater; the mix of
the two is a plumbing-and-cost optimisation on top of the thermodynamic gain that
every heater shares.

\paragraph{Back-pressure against condensing steam, the cogeneration trade.} A
steam plant asked to supply both power and process heat can bleed its heat from
the turbine two ways, and the choice sets how much electricity the heat costs.
Take steam raised at $6\,\text{MPa}/480\,\degC$, $h_3 = 3375\,\text{kJ}/\text{kg}$,
$s_3 = 6.79\,\text{kJ}/\text{kg}\cdot\text{K}$, serving a process that needs heat
at $150\,\degC$, which fixes a $0.5\,\text{MPa}$ delivery. A \engterm{back-pressure}
turbine expands to $0.5\,\text{MPa}$ and stops there, passing all its exhaust to
the process: the expansion to $h \approx 2790\,\text{kJ}/\text{kg}$ yields
$w_{\text{bp}} = 585\,\text{kJ}/\text{kg}$ of work, and every kilogram then
delivers $2790 - 640 = 2150\,\text{kJ}/\text{kg}$ of process heat as it condenses
in the process heater. The plant makes little power per unit steam but wastes
almost nothing, a first-law efficiency above $85\,\%$. A \engterm{condensing-extraction}
turbine instead bleeds only the steam the process needs at $0.5\,\text{MPa}$ and
expands the rest to a $10\,\text{kPa}$ condenser, reaching $h \approx
2160\,\text{kJ}/\text{kg}$ and yielding $w_{\text{cond}} = 1215\,\text{kJ}/\text{kg}$
on the condensing fraction, more than twice the back-pressure work per kilogram,
but the heat that fraction dumps to the condenser at $46\,\degC$ is nearly pure
waste. The trade is stark: the back-pressure machine sacrifices roughly
$1215 - 585 = 630\,\text{kJ}/\text{kg}$ of potential electricity on every
kilogram it sends to the process rather than to the condenser, which is the real,
non-zero cost of the ``free'' cogeneration heat the chapter's judgment exercise
warns against. Sizing the two turbine paths against the actual ratio of heat
demand to power demand, and against their prices, is the central design decision
of a cogeneration steam plant.

\paragraph{Second-law audit of the domestic refrigerator, component by component.}
The refrigerator teardown split the compressor input by first-law shares; the
exergy audit assigns each component the work potential it destroyed, which is the
quantity a redesign actually recovers. Take the $34\,\text{W}$ machine holding a
$-25\,\degC$ evaporator against a $40\,\degC$ condenser in a $22\,\degC$ kitchen,
dead state $T_0 = 295\,\si{\kelvin}$. The reversible minimum work to move the
$40\,\text{W}$ load across the device span is $\dot W_{\min} = \dot Q_c
(T_0 - T_{\text{evap}})/T_{\text{evap}}$ referred to the dead state, but read more
plainly as the load times the ratio of lost-to-useful, the machine's $34\,\text{W}$
input against a reversible $10.5\,\text{W}$ leaves $23.5\,\text{W}$ of exergy
destroyed to account for. The compressor and its motor, running at perhaps
$0.60$ combined isentropic-and-electrical efficiency, destroy the largest share,
near $9\,\text{W}$; the throttle, replacing a reversible expansion with an
isenthalpic drop, destroys $\dot X_{\text{throttle}} = T_0 \dot m (s_4 - s_3)$,
which for this deep-lift cycle is near $5\,\text{W}$; the condenser and evaporator
each destroy $T_0 \dot Q (1/T_{\text{fluid}} - 1/T_{\text{reservoir}})$ across
their finite pinches, near $4\,\text{W}$ and $3\,\text{W}$ respectively; and heat
leaking back through the cabinet walls carries off the rest. Ranked by exergy, the
compressor leads, the throttle and heat exchangers follow, and the wall leak,
which dominated the first-law load, destroys little exergy because it crosses a
small temperature gap. The audit therefore points a different finger than the
first-law load did: the load analysis said thicken the insulation to shrink the
duty, while the exergy analysis says the compressor and the pinch temperatures are
where the work is being destroyed once the duty is fixed. Both readings are
correct for their question, and a good redesign uses the first to cut the load and
the second to cut the loss per unit of the load that remains.

\paragraph{Heat-pump balance point, sized against a load line.} The balance point
the refrigeration section named qualitatively is fixed by two straight lines, and
computing it shows how heat-pump capacity is chosen. A house loses heat in
proportion to the indoor-outdoor difference, $\dot Q_{\text{load}} = UA
(T_{\text{in}} - T_{\text{out}})$; take $UA = 250\,\text{W}/\si{\kelvin}$ and
$T_{\text{in}} = 20\,\degC$, so at an outdoor $-10\,\degC$ the load is
$250 \times 30 = 7500\,\text{W}$. A heat pump's delivered capacity falls with
outdoor temperature roughly linearly over the range of interest; take a unit
rated $6000\,\text{W}$ at $8.3\,\degC$ falling to $3500\,\text{W}$ at $-8.3\,\degC$,
a slope of $2500\,\text{W}$ over $16.6\,\si{\kelvin}$, or $151\,\text{W}/\si{\kelvin}$.
Writing capacity as $\dot Q_{\text{cap}} = 3500 + 151(T_{\text{out}} + 8.3)$ and
load as $\dot Q_{\text{load}} = 250(20 - T_{\text{out}})$ and setting them equal,
$3500 + 151 T_{\text{out}} + 1253 = 5000 - 250 T_{\text{out}}$, so
$401 T_{\text{out}} = 247$, $T_{\text{out}} = 0.6\,\degC$: above about
$0.6\,\degC$ the pump carries the house alone, and below it the shortfall must
come from supplemental resistance heat at $\text{COP} = 1$. At the design
$-10\,\degC$ the load is $7500\,\text{W}$ and the pump delivers only
$3500 + 151 \times (-1.7) = 3243\,\text{W}$, so $4257\,\text{W}$ of resistance
backup runs, and that expensive fraction is what drags the seasonal COP below the
steady figure. Pushing the balance point colder means buying a larger unit whose
capacity line sits higher, which is why cold-climate heat pumps oversize the
compressor and add variable-speed capacity, trading capital cost and mild-weather
short-cycling for fewer hours on resistance heat.

\paragraph{HRSG pinch point and the steam pressure it selects.} The
heat-recovery steam generator of the combined-cycle plant cannot raise steam at an
arbitrary pressure, because the boiling temperature it fixes must stay below the
gas temperature at the point in the exchanger where the water begins to boil, the
\engterm{pinch point}. Take the topping-cycle exhaust entering at $527\,\degC$
($800\,\si{\kelvin}$) and cooling toward the stack, and ask what steam pressure a
$15\,\si{\kelvin}$ pinch permits. The gas gives up its heat first to superheat and
boil the water, then to preheat it; the tightest gap is where the gas meets the
saturated water at the start of boiling, so the boiling temperature can be no
higher than the local gas temperature less the pinch. If the gas at the
evaporator inlet is near $330\,\degC$ after giving up its superheat duty, a
$15\,\si{\kelvin}$ pinch caps the boiling temperature at $315\,\degC$, whose
saturation pressure for water is about $10.6\,\text{MPa}$, which is why the worked
plant chose a $10\,\text{MPa}$ boiler. Raising the steam pressure to lift the
Rankine efficiency raises the boiling temperature and so squeezes the pinch, and
past the point where the boiling temperature reaches the gas temperature less the
pinch the exchanger stalls: no more heat can cross. The resolution large plants
adopt is multi-pressure heat recovery, raising steam at two or three pressures so
that a high-pressure circuit takes the hottest gas and a low-pressure circuit
mops up the cooler gas near the stack, each at its own pinch, which recovers more
of the exhaust than any single pressure could. The pinch point is the single
constraint that ties the HRSG steam pressure to the gas-turbine exhaust
temperature, and it is why a hotter topping cycle both raises the bottoming
cycle's efficiency and relaxes its pinch at once.

\paragraph{Reversed Brayton cooling with real components.} The aircraft air-cycle
example ran the reversed Brayton cycle with ideal expansion; a ground gas-cycle
cooler with real components shows how quickly the modest ideal COP erodes. Take
air drawn at $T_1 = 300\,\si{\kelvin}$, compressed at a pressure ratio $r_p = 3$
through a compressor of isentropic efficiency $\eta_c = 0.80$, cooled in a heat
exchanger back to $T_3 = 310\,\si{\kelvin}$ against ambient, then expanded through
a turbine of isentropic efficiency $\eta_t = 0.82$ to provide cooling. The
isentropic compressor exit is $T_{2s} = 300 \times 3^{0.286} = 411\,\si{\kelvin}$,
so the real exit is $T_2 = 300 + (411 - 300)/0.80 = 439\,\si{\kelvin}$. The
isentropic turbine exit is $T_{4s} = 310 \times 3^{-0.286} = 226\,\si{\kelvin}$,
so the real exit is $T_4 = 310 - 0.82 \times (310 - 226) = 241\,\si{\kelvin}$,
about $-32\,\degC$, the cold air delivered to the load. The cooling effect is
$q_c = c_p(T_1' - T_4)$ where the cold air warms back to the load return
temperature; taking the load return at $T_3 = 310\,\si{\kelvin}$ gives $q_c =
1.005 \times (310 - 241) = 69\,\text{kJ}/\text{kg}$. The net work is the
compressor work less the turbine work recovered, $w_{\text{net}} = c_p[(T_2 - T_1)
- (T_3 - T_4)] = 1.005 \times [(439 - 300) - (310 - 241)] = 1.005 \times (139 -
69) = 70\,\text{kJ}/\text{kg}$, so the COP is $69/70 = 0.99$, against the ideal
reversed-Brayton COP $1/(3^{0.286} - 1) = 2.7$ the section quoted. Two real
component efficiencies of about $0.80$ cut the COP from $2.7$ to $1.0$, a far
steeper collapse than the same inefficiencies inflict on a vapour-compression
cycle, because the reversed Brayton cycle, like its engine cousin, carries a heavy
back-work ratio: the turbine work is most of the compressor work, so a loss in
each machine bites twice. That sensitivity, not the ideal COP, is the honest
reason the gas cycle is reserved for the cryogenic and aerospace duties where no
vapour cycle can serve.

\paragraph{Combined-cycle exergy split, stack against condenser.} The
combined-cycle case study asserted that the visibly hot stack destroys little work
potential while the invisible combustion step destroys the most; a short exergy
calculation on the two low-temperature rejections makes the claim quantitative.
The stack rejects heat at $390\,\si{\kelvin}$ ($117\,\degC$) to a dead state at
$T_0 = 295\,\si{\kelvin}$, so each joule leaving the stack carries an exergy of
$1 - 295/390 = 0.24$ joule. The condenser rejects heat at $315\,\si{\kelvin}$
($42\,\degC$, the $8\,\text{kPa}$ saturation temperature), so each joule there
carries only $1 - 295/315 = 0.064$ joule of exergy. Per kilogram of topping-cycle
air the stack carries away $c_p(390 - 295) = 1.005 \times 95 = 95\,\text{kJ}$
above the dead state, of which $0.24 \times 95 = 23\,\text{kJ}$ is exergy; the
condenser rejects the bottoming cycle's roughly $264\,\text{kJ}$ of heat, of which
only $0.064 \times 264 = 17\,\text{kJ}$ is exergy. The two large first-law losses,
$95$ and $264\,\text{kJ}$, carry only $23$ and $17\,\text{kJ}$ of work potential
between them, a combined $40\,\text{kJ}$ against the $887\,\text{kJ}$ of fuel
exergy, under five percent. The combustion step, by contrast, destroys near
$28\,\%$ of the fuel exergy by releasing chemical energy as heat across the gap
between the flame near $2000\,\si{\kelvin}$ and the working gas. The arithmetic is
the reason the exergy waterfall of figure \ref{fig:vol04ch03:exergy-destruction}
points a redesign at the combustor and the heat-addition path rather than at the
stack the first law makes look wasteful: a joule rejected cold is a joule that had
almost no work left in it to lose.

\paragraph{Turbine-blade cooling and its optimum inlet temperature.} The Brayton
examples raised the turbine inlet temperature freely, but past the metal's limit a
gas turbine survives only by bleeding cool air from the compressor to film-cool and
internally cool the first turbine stages, and that bleed is a debit the cycle
carries. Take the $r_p = 18$ topping cycle of the combined-cycle study, inlet
$T_1 = 288\,\si{\kelvin}$, compressor exit $T_2 = 717\,\si{\kelvin}$, and let the
uncooled metal limit be $1300\,\si{\kelvin}$. Running the inlet at $T_3 =
1600\,\si{\kelvin}$ demands cooling, and a rough rule sizes the cooling-air
fraction as proportional to the excess above the metal limit: taking a fraction
$f = 0.05$ of the compressor flow per $150\,\si{\kelvin}$ of excess gives $f =
0.05 \times (1600 - 1300)/150 = 0.10$, a tenth of the air routed around the
combustor. That tenth is compressed (its share of the $431\,\text{kJ}/\text{kg}$
compressor work is spent) but adds no combustion heat, and when it dumps into the
gas path it dilutes and cools the working gas, so the effective turbine inlet
temperature falls below the nominal $1600\,\si{\kelvin}$ toward a mixed
$1600 - 0.10 \times (1600 - 717) = 1512\,\si{\kelvin}$. The turbine work per
kilogram of total air falls accordingly, while the combustor still burns fuel only
for the uncooled ninety percent. Sweeping $T_3$ traces figure
\ref{fig:vol04ch03:blade-cooling-penalty}: the net work rises with inlet
temperature at first, because the hotter gas returns more than the cooling air
costs, then bends over as the cooling fraction climbs faster than the temperature
buys, reaching an optimum inlet temperature that a real cooled turbine sits near.
The reading is that the inlet-temperature ceiling is not the metal melting point
but the point where the marginal cooling air costs more work than the marginal
degree of gas temperature returns, and every advance in cooling technology and
thermal-barrier coating moves that economic optimum to a higher temperature rather
than removing the penalty.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-blade-cooling-penalty.tex}

\paragraph{Combustor energy balance and the fuel-air ratio.} The air-standard
cycles modelled heat addition as a number $q_h$; a real combustor fixes that number
through the fuel it burns, and a short energy balance ties the turbine inlet
temperature to the fuel-air ratio the machine actually meters. Take the
combined-cycle topping combustor, air entering at $T_2 = 717\,\si{\kelvin}$ and
leaving at $T_3 = 1600\,\si{\kelvin}$, burning natural gas of lower heating value
$50\,\text{MJ}/\text{kg}$. Per kilogram of air the combustor must add $q_h =
c_p(T_3 - T_2) = 1.005 \times 883 = 887\,\text{kJ}/\text{kg}$, and neglecting the
fuel's own thermal mass and taking a combustion efficiency of $0.99$, the
fuel-air ratio follows from $f\,\eta_{\text{comb}}\,\text{LHV} = q_h(1 + f)
\approx q_h$, so $f = q_h/(\eta_{\text{comb}}\,\text{LHV}) = 887/(0.99 \times
50{,}000) = 0.0179$, about one part fuel to fifty-six parts air by mass. That is far
leaner than the stoichiometric ratio for natural gas, near $1{:}17$, so the gas
turbine runs at roughly a third of stoichiometric, three times the air a complete
burn needs. The excess air is not waste; it is the coolant that holds the turbine
inlet temperature down to what the blades tolerate, since a stoichiometric flame
would sit near $2300\,\si{\kelvin}$, far above any blade limit. The lean operation
the cycle demands for blade survival is the same lean operation that suppresses the
thermal nitric-oxide formation the flame temperature drives, which is why the
combustor's air split serves emissions and metallurgy at once. The fuel-air ratio,
not an abstract $q_h$, is the quantity a control system meters, and it falls
directly out of the inlet-temperature target the blade metallurgy sets.

\paragraph{Lorenz bound against a gliding heat sink.} The Carnot and Curzon-Ahlborn
bounds assumed infinite reservoirs at fixed temperature; a refrigerator rejecting
heat to a finite cooling-water stream faces a stricter bound, because the water
warms as it carries the heat away and the effective sink temperature is not its
inlet value. The \engterm{Lorenz cycle} is the reversible cycle that exchanges heat
with such sensible streams, and its bound uses the log-mean temperature of each
stream in place of a single reservoir temperature. Take a chiller holding a cold
load at $T_c = 275\,\si{\kelvin}$ and rejecting to cooling water that enters at
$305\,\si{\kelvin}$ and leaves $20\,\si{\kelvin}$ warmer at $325\,\si{\kelvin}$.
Reading the reversible COP from the water inlet gives an optimistic
$T_c/(T_{h,\text{in}} - T_c) = 275/(305 - 275) = 9.2$, but the honest bound uses
the water's log-mean temperature
\[
T_{h,\text{lm}} = \frac{T_{h,\text{out}} - T_{h,\text{in}}}
{\ln(T_{h,\text{out}}/T_{h,\text{in}})}
= \frac{20}{\ln(325/305)} = 315\,\si{\kelvin},
\]
so the Lorenz COP is $T_c/(T_{h,\text{lm}} - T_c) = 275/(315 - 275) = 6.9$, a
quarter below the inlet figure (figure \ref{fig:vol04ch03:lorenz-bound}). The gap
widens as the glide widens, which is the quantitative reason a chiller sized against
its cooling-water inlet temperature underperforms in service once the water has
warmed across the condenser, and why undersizing the cooling-water flow, which
forces a wider glide, hurts the COP twice: once through the higher mean rejection
temperature and once through the pinch it opens in the condenser. The Lorenz bound
is the refrigeration reading of the same average-temperature accounting the
gliding-source power cycles turned on, and it is the correct yardstick whenever a
reservoir is a stream rather than an ocean.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-lorenz-bound.tex}

\paragraph{Adiabatic compressed-air energy storage, round trip.} A storage plant
is two machines back to back, a compressor charging a store and an expander
discharging it, and its figure of merit is the electricity-to-electricity round
trip, which the heat of compression governs. Take air compressed from
$100\,\text{kPa}$ and $300\,\si{\kelvin}$ to a cavern at $7\,\text{MPa}$, a
pressure ratio of $70$. A single-stage isentropic compression would reach
$T_2 = 300 \times 70^{0.286} = 300 \times 3.44 = 1032\,\si{\kelvin}$, too hot to
store, so real plants compress in stages. The heat of compression, $c_p(T_2 - T_1)
= 1.005 \times 732 = 736\,\text{kJ}/\text{kg}$, is the crux: a diabatic plant cools
the air between stages and after the last, discards that $736\,\text{kJ}/\text{kg}$
to the atmosphere, and then, on discharge, must reheat the cold cavern air by
burning fuel before it expands, because expanding cold air would reach cryogenic
temperatures and deliver little work. That plant is really a gas turbine fed by
stored air, and its electricity-only round trip, charged against the reheat fuel,
is modest. An adiabatic plant instead stores the $736\,\text{kJ}/\text{kg}$ of
compression heat in a thermal reservoir, a bed of rock or a molten-salt loop, and
returns it to the air on discharge, so no fuel burns and the expansion recovers the
work the compression stored. Because the compression heat is exactly the exergy the
expansion needs back, the round-trip efficiency rises with the fraction of that heat
the store returns (figure \ref{fig:vol04ch03:caes-roundtrip}): a diabatic plant
sits near forty percent, an adiabatic plant with a good thermal store reaches
seventy. The lesson is that in any storage device built from a heat cycle, the
irreversibility to hunt is the heat thrown away between the charge and discharge
halves, and the design move that pays is to store that heat rather than to improve
either machine, the same store-the-regenerated-heat logic the Stirling and Ericsson
cycles turned on, applied across a charge-discharge gap in time rather than around
a loop in space.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-caes-roundtrip.tex}

\paragraph{Reheat-pressure optimum by maximising the average heat-addition
temperature.} The reheat worked example placed the reheat pressure near a quarter
of the boiler pressure by assertion; the optimum follows from a short argument that
the reheat pressure maximises the average temperature at which the second batch of
heat enters. Reheating adds heat $q_{\text{rh}} = h_5 - h_4$ at the reheat pressure,
raising the steam from the cold end of the high-pressure expansion back to the
turbine inlet temperature. Set the reheat pressure too high and the high-pressure
expansion is short, so little steam reaches the reheater and the reheat adds almost
no heat; set it too low and the reheater takes the steam in already cool and adds
its heat at a temperature barely above the low-pressure turbine's, a poor place to
add heat. Between these the average temperature of the reheat heat addition passes
through a maximum. For the $15\,\text{MPa}/550\,\degC$ boiler of the reheat example,
sweeping the reheat pressure shows the average heat-addition temperature peaking
broadly between about a fifth and a third of the boiler pressure; at
$2\,\text{MPa}$, the example's choice, it sits near that peak, which is why the
efficiency was insensitive to moving the reheat pressure a little either way. The
same broad, flat optimum appears whenever a heat input is staged to raise its
average temperature, and its flatness is a gift to the designer, who can place the
reheat tap where the turbine casing and the pipe runs are convenient without paying
much efficiency for the freedom. The reheat pressure is chosen, in the end, by the
exit-quality constraint of the wet-steam subsection as much as by this efficiency
optimum, and the two considerations happen to point at the same neighbourhood.

\paragraph{Evaporative cooling, cold from a water spray.} Not every cooling duty
needs a compressor. Where the air is dry, spraying water into it cools the air as
the water evaporates, drawing its latent heat from the airstream, and the
thermodynamic bound on that cooling is the wet-bulb temperature. Take outdoor air
at $38\,\degC$ dry-bulb and $15\,\%$ relative humidity, a hot dry climate. The
adiabatic saturation of that air, following a line of constant enthalpy on the
psychrometric chart, brings it toward a wet-bulb temperature near $20\,\degC$, so a
direct evaporative cooler can in principle deliver air at about $20\,\degC$, an
$18\,\si{\kelvin}$ drop, for only the small fan and pump power that move the air and
the water. The cooling per kilogram of air is $c_p \Delta T = 1.005 \times 18 =
18\,\text{kJ}/\text{kg}$, delivered for a compressor work of nearly zero, so
the effective coefficient of performance measured against the tiny electrical input
runs into the tens, far above any vapour-compression figure. The catch is written
into the same psychrometric line: the cooling stops at the wet-bulb temperature and
adds moisture to the air, so a humid climate, where the wet-bulb sits close to the
dry-bulb, gets almost no cooling, and even a dry climate cannot push the air below
its wet-bulb no matter how much water is sprayed. The vapour-compression machine
buys its independence from humidity and its ability to cool below the wet-bulb with
the compressor work the evaporative cooler avoids, which is the trade between the two
technologies. The evaporative cooler is the clearest case in the chapter that the
cheapest way to move heat is sometimes to change phase directly in the load rather
than to run a cycle at all, and it explains why the technology dominates in hot dry
regions and vanishes in humid ones.

The chapter's separate cycles meet in one machine when an engineer must
size and rate a combined-cycle power plant. We carry a single fuel input
through a Brayton topping cycle, a heat-recovery steam generator, and a
Rankine bottoming cycle, to a net plant efficiency and an electrical
rating, then read the result against the second law. Every section of the
chapter appears once: the Brayton pressure ratio and turbine inlet
temperature, the back-work ratio, the heat-recovery temperature window,
the Rankine boiler and condenser conditions, and the combined-cycle
efficiency formula of section 3.6. The plant is the one whose energy flow
figure \ref{fig:vol04ch03:combined-cycle-flow} sketched; here we put the
numbers on it.

\subsection{The topping cycle}

The gas turbine draws air at $T_1 = 288\,\si{\kelvin}$ ($15\,\degC$, the
standard rating condition) and $101\,\text{kPa}$, compresses it at a
pressure ratio $r_p = 18$ through a compressor of isentropic efficiency
$\eta_c = 0.86$, burns natural gas to a turbine inlet temperature $T_3 =
1600\,\si{\kelvin}$, and expands through a turbine of isentropic
efficiency $\eta_t = 0.89$ back to atmospheric pressure. With $\gamma =
1.4$ and exponent $0.286$, the isentropic compressor exit is $T_{2s} =
288 \times 18^{0.286} = 657\,\si{\kelvin}$, so the real exit is $T_2 =
288 + (657 - 288)/0.86 = 717\,\si{\kelvin}$. The isentropic turbine exit
is $T_{4s} = 1600 \times 18^{-0.286} = 701\,\si{\kelvin}$, so the real
exit is $T_4 = 1600 - 0.89 \times (1600 - 701) = 800\,\si{\kelvin}$, that
is $527\,\degC$, the temperature that sets the heat available to the
bottoming cycle.

Per kilogram of air, with $c_p = 1.005\,\text{kJ}/\text{kg}\cdot\text{K}$:
the turbine work is $c_p(T_3 - T_4) = 1.005 \times 800 = 804\,\text{kJ}/
\text{kg}$, the compressor work is $c_p(T_2 - T_1) = 1.005 \times 429 =
431\,\text{kJ}/\text{kg}$, and the net gas-turbine work is $373\,\text{kJ}
/\text{kg}$. The back-work ratio is $431/804 = 0.54$, on the high side
because the pressure ratio is high; over half the turbine output runs the
compressor. The heat added in the combustor is $c_p(T_3 - T_2) = 1.005
\times 883 = 887\,\text{kJ}/\text{kg}$, so the gas-turbine thermal
efficiency standing alone is $\eta_{\text{GT}} = 373/887 = 0.421$, about
$42\,\%$. The exhaust carries the balance, $887 - 373 = 514\,\text{kJ}/
\text{kg}$, away at $527\,\degC$, and that hot stream is the resource the
rest of the plant exists to harvest.

\subsection{The heat-recovery steam generator}

The exhaust at $800\,\si{\kelvin}$ cannot be cooled all the way to ambient
in the steam generator: the stack must stay above the water dew point to
avoid acid condensation, and the steam it raises cannot be hotter than
the gas that heats it. We cool the exhaust to a stack temperature of
$390\,\si{\kelvin}$ ($117\,\degC$), recovering $c_p(800 - 390) = 1.005
\times 410 = 412\,\text{kJ}$ per kilogram of air, which is $412/514 =
0.80$ of the exhaust heat above ambient. That recovered $412\,\text{kJ}/
\text{kg-air}$ is the heat input to the Rankine cycle. The temperature
window the generator works across, from the $527\,\degC$ gas inlet down
to the $117\,\degC$ stack, is the same window figure
\ref{fig:vol04ch03:combined-cycle-ts} marks between its dashed lines; the
steam the generator raises must thread that window, which fixes the steam
pressure and the achievable superheat.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-combined-cycle-ts.tex}

\subsection{The bottoming cycle}

The steam cycle takes that heat at a boiler pressure of $10\,\text{MPa}$,
superheated to $500\,\degC$ (held below the $527\,\degC$ gas inlet by the
pinch the heat exchanger needs), and rejects to a condenser at
$8\,\text{kPa}$, the saturation pressure for the available cooling water.
From the worked Rankine arithmetic of this chapter, a $10\,\text{MPa}/
500\,\degC$ boiler against an $8\,\text{kPa}$ condenser, with a turbine
isentropic efficiency of $0.88$, gives a steam-cycle thermal efficiency
near $\eta_{\text{ST}} = 0.36$. The bottoming cycle therefore converts
$0.36 \times 412 = 148\,\text{kJ}$ of the recovered heat to work, per
kilogram of the topping cycle's air.

\subsection{The plant rating}

The two cycles together deliver $373 + 148 = 521\,\text{kJ}$ of work per
kilogram of air, against the $887\,\text{kJ}$ of fuel heat the combustor
supplied. The combined-cycle thermal efficiency is
\[
\eta_{cc} = \frac{521}{887} = 0.587,
\]
about $59\,\%$, which the section 3.6 formula reproduces:
$\eta_{cc} = \eta_{\text{GT}} + \eta_{\text{HRSG}}(1 - \eta_{\text{GT}})
\eta_{\text{ST}} = 0.421 + 0.80 \times 0.579 \times 0.36 = 0.421 + 0.167
= 0.588$. The gas turbine alone reached $42\,\%$; reusing its exhaust
lifted the plant to $59\,\%$, the difference being the $148\,\text{kJ}$
the bottoming cycle squeezed from heat the single cycle threw away.

To reach a $400\,\text{MW}$ net electrical rating, allow a generator and
mechanical efficiency of $0.985$ on each shaft and a plant auxiliary load
(pumps, fans, controls) of about $2\,\%$ of gross. The net specific work
is then roughly $0.965 \times 521 = 503\,\text{kJ}/\text{kg-air}$, so the
air mass flow is
\[
\dot m_{\text{air}} = \frac{400 \times 10^3\,\si{\kilo\watt}}
{503\,\text{kJ}/\text{kg}} = 795\,\text{kg}/\text{s},
\]
a flow a single large heavy-duty gas turbine plus its steam train can
supply. The fuel power follows from the efficiency: $\dot Q_{\text{fuel}}
= 400/0.587 \times (503/521) = 658\,\text{MW}$ of natural gas, about
$13\,\text{kg}/\text{s}$ at a heating value of $50\,\text{MJ}/\text{kg}$.
A reader checking these against a published rating plate will find them in
the right neighbourhood; the published machines reach a point or two
higher through multi-pressure heat recovery and reheat in the steam
train, refinements the next subsection reads off the second law.

\subsection{Where the work potential went}

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-exergy-destruction.tex}

The first law says $41\,\%$ of the fuel energy left unconverted, split
between the stack and the condenser. That accounting is silent on which
loss was avoidable. The exergy accounting of figure
\ref{fig:vol04ch03:exergy-destruction} answers the question the plant
manager actually has. Of the fuel's work potential, the combustion step
destroys the largest single share, near $28\,\%$, by releasing chemical
energy as heat across an enormous temperature gap between the flame and
the working gas; no turbine in the plant ever sees that lost potential.
The compressor and turbine irreversibilities, the ones the component
efficiencies $\eta_c$ and $\eta_t$ measured, account for a few percent
each, and the steam train a few more. The stack and the condenser, which
carry most of the first-law energy loss, carry little exergy, because
they reject heat at low temperature where each joule holds little work
value. The design moral is that the visibly hot stack is not where the
plant loses its capacity to do work; the combustor and the heat-addition
path are. That is the reasoning behind every refinement that raises the
turbine inlet temperature or recovers exhaust heat at multiple pressures:
each one narrows a temperature gap across which work potential was being
spent for nothing.

\section{An end-to-end case study: a compact-car gasoline engine}\pathtag{standard}

The combined-cycle plant carried one fuel input through three cycles to a
plant rating; the reciprocating engine is the other end of the scale, and
carrying a single road condition from the air-standard cycle through the
indicated work, the friction and pumping losses, the brake output, and the
fuel consumption ties the Otto-cycle arithmetic of section 3.4 to a number a
driver can read off a fuel gauge. We take a naturally aspirated four-cylinder
gasoline engine of $1.6\,\text{L}$ displacement, geometric compression ratio
$r = 10.5$, driving a $1400\,\text{kg}$ car at a steady
$100\,\text{km}/\text{h}$ cruise, and follow the energy from the air-standard
ceiling down to the road.

\subsection{The air-standard ceiling}

The ideal Otto efficiency for $r = 10.5$ and $\gamma = 1.4$ is
$\eta_O = 1 - 10.5^{-0.4} = 1 - 0.392 = 0.608$, about $61\,\%$. This is the
ceiling the rest of the analysis falls below, and naming it first fixes how
much of the loss is the cycle's geometry and how much is the hardware. A more
honest air-standard estimate uses a ratio of specific heats lower than the
cold-air $1.4$, because the hot combustion products have more active molecular
modes; taking $\gamma = 1.30$ for the high-temperature portion gives
$1 - 10.5^{-0.30} = 1 - 0.494 = 0.506$, about $51\,\%$. The drop from $61\,\%$
to $51\,\%$ on a single more realistic property is the first warning that the
cold-air-standard number flatters the cycle. A fuel-air cycle analysis, which
also accounts for the charge's dilution by residual gas and the temperature
dependence of the specific heats through the burn, lands near $0.47$ for this
compression ratio. We carry $\eta_{\text{fa}} = 0.47$ as the realistic
upper bound on the indicated thermal efficiency.

\subsection{Indicated work and the road demand}

The road load sets what the engine must actually deliver. Rolling resistance
at $C_r = 0.010$ on a $1400\,\text{kg}$ car is $F_{\text{roll}} = 0.010 \times
1400 \times 9.81 = 137\,\text{N}$; aerodynamic drag with $C_d A = 0.62\,
\text{m}^2$ at air density $1.20\,\text{kg}/\text{m}^3$ and
$v = 27.8\,\text{m}/\text{s}$ is $F_{\text{drag}} = 0.5 \times 1.20 \times
0.62 \times 27.8^2 = 288\,\text{N}$. The total road force is $425\,\text{N}$,
so the road power is $425 \times 27.8 = 11.8\,\text{kW}$. Allowing a
drivetrain efficiency of $0.88$ from the crankshaft to the wheels, the engine
must produce a brake power of $11.8/0.88 = 13.4\,\text{kW}$ at the wheels'
demand, plus an accessory load (alternator, water pump, power steering) of
about $0.8\,\text{kW}$, for a brake output near $14.2\,\text{kW}$ at this
cruise.

The engine reaches that brake output at a part-throttle operating point well
below its rated power, and that is where the efficiency story turns. At
$100\,\text{km}/\text{h}$ in top gear the engine turns near
$2200\,\text{rpm}$, where its wide-open-throttle torque would be about
$130\,\text{N}\cdot\text{m}$ (figure \ref{fig:vol04ch03:engine-power-bsfc}),
hence a wide-open brake power near $2\pi \times (2200/60) \times 130 =
30\,\text{kW}$. Delivering only $14.2\,\text{kW}$ means the throttle is nearly
half closed, and the partly closed throttle is a pump the engine drives
against its own intake: the \engterm{pumping loss}. The pumping mean effective
pressure at this part load is roughly $0.4\,\text{bar}$, against an indicated
mean effective pressure near $4\,\text{bar}$ at this load, so pumping alone
consumes about $10\,\%$ of the indicated work before friction is counted.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-engine-power-bsfc.tex}

\subsection{From indicated to brake}

The indicated work is what the gas does on the piston; the brake work is what
leaves the crankshaft, and the gap is the \engterm{mechanical efficiency}. The
friction mean effective pressure of a warm engine at $2200\,\text{rpm}$ is
about $1.0\,\text{bar}$, and the pumping mean effective pressure is the
$0.4\,\text{bar}$ above, so the total loss is $1.4\,\text{bar}$ against a gross
indicated mean effective pressure near $4.4\,\text{bar}$. The mechanical
efficiency is therefore $\eta_{\text{mech}} = (4.4 - 1.4)/4.4 = 0.68$ at this
part-load point, far below the $0.85$ to $0.90$ a wide-open-throttle point
reaches, because the losses are nearly fixed in absolute terms while the
indicated work has shrunk with the throttling. The brake thermal efficiency is
the product of the indicated thermal efficiency and the mechanical efficiency.
Taking an indicated thermal efficiency of $0.38$ at this part-load point (below
the $0.47$ ceiling because of the lower load and the spark timing the engine
runs for smoothness rather than peak efficiency), the brake thermal efficiency
is $\eta_b = 0.38 \times 0.68 = 0.26$, about $26\,\%$.

\subsection{Brake specific fuel consumption and the tank}

The brake specific fuel consumption converts the brake efficiency into the
number a driver pays. With a gasoline lower heating value of
$43.5\,\text{MJ}/\text{kg}$,
\[
\text{BSFC} = \frac{1}{\eta_b \, \text{LHV}} =
\frac{1}{0.26 \times 43.5\,\text{MJ}/\text{kg}} =
8.8 \times 10^{-8}\,\text{kg}/\text{J} = 318\,\text{g}/\text{kW}\cdot\text{h},
\]
on the high side of the engine's map (figure
\ref{fig:vol04ch03:engine-power-bsfc} shows the best BSFC near
$243\,\text{g}/\text{kW}\cdot\text{h}$ at the torque peak, reached only at high
load). The fuel power at the $14.2\,\text{kW}$ brake output is
$\dot Q_{\text{fuel}} = 14.2/0.26 = 54.6\,\text{kW}$, so the fuel mass rate is
$54.6/43.5 = 1.26\,\text{g}/\text{s}$, or with a gasoline density of
$0.745\,\text{kg}/\text{L}$, a volume rate of $1.68\,\text{mL}/\text{s}$. Over
the $36\,\text{s}$ the car takes to travel one kilometre at
$100\,\text{km}/\text{h}$, that is $60\,\text{mL}/\text{km}$, which is
$6.0\,\text{L}/100\,\text{km}$, squarely in the range a real compact car posts
at steady highway cruise.

\subsection{Where the fuel went}

Of the $54.6\,\text{kW}$ of fuel power, $14.2\,\text{kW}$ ($26\,\%$) left as
brake work and $11.8\,\text{kW}$ ($22\,\%$) reached the road; the
$2.4\,\text{kW}$ difference is drivetrain and accessory loss. The remaining
$40.4\,\text{kW}$ ($74\,\%$) divides into roughly equal thirds: exhaust enthalpy
carried out the tailpipe (about $30\,\%$ of the fuel), heat to the coolant and
oil (about $28\,\%$), and the combined friction, pumping, and incomplete
combustion losses inside the engine (about $16\,\%$). The single largest stream
is the exhaust, which is why exhaust-gas turbocharging and, in the combined
heat of the cycle, the exhaust-driven Rankine bottoming systems some heavy
trucks now carry, target that stream first. The contrast with the
combined-cycle plant is the whole point of placing the two case studies
together: the stationary plant reaches $59\,\%$ because it has the size and the
steady load to chase the exhaust heat through a second cycle, while the car
engine surrenders that heat because no bottoming cycle small and light enough to
ride along has yet been worth its weight and cost. The same second law governs
both; the scale and the duty decide how much of the bound each machine can
afford to approach.

\section{An end-to-end case study: an organic Rankine cycle on industrial waste heat}\pathtag{standard}

The combined-cycle plant and the car engine both burned fuel; a third class of
machine burns nothing and lives on heat that another process has already thrown
away. A cement kiln, a glass furnace, a gas-engine generator, a steel reheat
furnace, each rejects a large stream of exhaust at a few hundred Celsius, hot
enough to carry real work potential yet too cool for a steam cycle to harvest
economically. The organic Rankine cycle is the standard answer: a Rankine cycle
with an organic working fluid chosen so that its saturation curve fits the modest
temperatures of the waste stream. We carry one such installation from the
exhaust duct to a net electrical rating, and the arithmetic ties the Rankine
states of section 3.2 to a working-fluid choice the steam cycle never has to make.

\subsection{The waste-heat source and the working fluid}

Take the exhaust of a $4\,\text{MW}$ reciprocating gas-engine generator at a
landfill-gas site: $11\,\text{kg}/\text{s}$ of exhaust leaving the engine at
$450\,\degC$, which a heat-recovery exchanger can cool to about $150\,\degC$
before the acid dew point and the pinch against the working fluid stop it. The
heat available is $\dot Q_{\text{src}} = \dot m_g c_{p,g}(450 - 150) = 11 \times
1.10 \times 300 = 3630\,\text{kW}$, taking the exhaust specific heat as
$1.10\,\text{kJ}/\text{kg}\cdot\text{K}$. That $3.63\,\text{MW}$ of heat at a
mean temperature near $300\,\degC$ is the resource.

Water would be a poor working fluid here. Raising steam at a useful pressure
needs a high temperature, the expansion would end deep in the wet region without
heavy superheat, and the volume flows at a low condenser pressure would force an
oversized turbine. An organic fluid sidesteps all three. We select a dry fluid,
one whose saturated-vapour line leans to the right on the $T$-$s$ plane
(figure \ref{fig:vol04ch03:orc-ts}), so that expansion from saturated vapour ends
dry with no superheat required. For a source of this temperature a hydrocarbon or
siloxane with a critical temperature comfortably above the evaporator temperature
fits; figure \ref{fig:vol04ch03:orc-fluid-selection} sets out the efficiency
ceiling each candidate allows. We carry toluene, critical temperature
$319\,\degC$, evaporating at $150\,\degC$ and condensing at $35\,\degC$ against
the site's cooling water.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-orc-ts.tex}

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-orc-fluid-selection.tex}

\subsection{The cycle states}

Toluene evaporating at $150\,\degC$ sits at about $0.50\,\text{MPa}$; condensing
at $35\,\degC$ sits near $0.008\,\text{MPa}$. Following the four Rankine states
with representative toluene enthalpies: the condenser delivers saturated liquid
at $h_1 = 12\,\text{kJ}/\text{kg}$ (taking the liquid enthalpy reference at the
condenser), the feed pump raises it across the $0.49\,\text{MPa}$ rise with
$v_f \approx 0.00115\,\text{m}^3/\text{kg}$, so $w_p = v_f \Delta p = 0.00115
\times 492 = 0.57\,\text{kJ}/\text{kg}$ and $h_2 = 12.6\,\text{kJ}/\text{kg}$.
The evaporator heats and boils the liquid to saturated vapour at $150\,\degC$,
$h_3 = 423\,\text{kJ}/\text{kg}$, so the heat added is $q_h = h_3 - h_2 =
410\,\text{kJ}/\text{kg}$. The turbine expands the dry vapour to the condenser
pressure; an isentropic expansion would reach $h_{4s} = 348\,\text{kJ}/\text{kg}$,
and a turbine of isentropic efficiency $\eta_t = 0.80$ delivers $h_3 - h_4 =
\eta_t(h_3 - h_{4s}) = 0.80 \times 75 = 60\,\text{kJ}/\text{kg}$, leaving the
exit dry to the right of the dome, where no blade erosion threatens.

\subsection{Net power and efficiency}

The net specific work is $w_{\text{net}} = (h_3 - h_4) - w_p = 60 - 0.57 =
59.4\,\text{kJ}/\text{kg}$, so the cycle thermal efficiency is $\eta_{\text{ORC}}
= w_{\text{net}}/q_h = 59.4/410 = 0.145$, about $15\,\%$. The pump work is a
hundredth of the turbine work, against the steam cycle's already small pump
fraction; the dense organic liquid and the modest pressure rise make the back-work
ratio negligible, one of the quiet advantages of the organic fluid. The working
fluid mass flow follows from the heat the source delivers,
\[
\dot m_{\text{wf}} = \frac{\dot Q_{\text{src}}}{q_h} =
\frac{3630\,\si{\kilo\watt}}{410\,\text{kJ}/\text{kg}} = 8.9\,\text{kg}/\text{s},
\]
and the gross turbine power is $\dot m_{\text{wf}}(h_3 - h_4) = 8.9 \times 60 =
531\,\text{kW}$. Subtracting the pump power $\dot m_{\text{wf}} w_p = 8.9 \times
0.57 = 5\,\text{kW}$ and allowing a generator and mechanical efficiency of
$0.96$ gives a net electrical output near $(531 - 5) \times 0.96 = 505\,\text{kW}$.

That half-megawatt is added to the $4\,\text{MW}$ the gas engine already produces,
for no extra fuel: the ORC lifts the site's electrical output by about $13\,\%$
while burning nothing. Read against its own resource the cycle reaches only
$15\,\%$, far below the combined-cycle plant's $59\,\%$, but the comparison is
the wrong one. The combined-cycle plant chose its own high source temperature by
burning fuel; the ORC took a source it did not pay for, at a temperature fixed by
someone else's exhaust, and any work it extracts is work that was leaving the site
as warm air. The right yardstick is the second-law efficiency against the source:
the Carnot bound between the $300\,\degC$ mean source and the $35\,\degC$ sink is
$1 - 308/573 = 0.46$, so the cycle's $15\,\%$ is a second-law efficiency near
$0.32$, typical for a small ORC and limited mostly by the $0.80$ turbine and the
temperature glide between the cooling exhaust and the constant-temperature
boiling toluene. Multi-pressure evaporation or a fluid with a better temperature
match to the glide would narrow that gap, the same average-temperature-of-heat-addition
lever the steam Rankine refinements pulled, applied to a cycle a hundred degrees
cooler.

\begin{mastery}
The choice between a wet, dry, and isentropic working fluid is read off the slope
of the saturated-vapour line on the $T$-$s$ plane, and it decides whether a cycle
needs superheat. For a fluid the saturated-vapour line slopes back over the dome
(a negative $ds/dT$ along saturation), so an isentropic expansion that starts on
that line crosses into the two-phase region and ends wet; the cycle must superheat
the vapour before the turbine to keep the exit dry, which is why every steam plant
superheats. A \engterm{dry} fluid has a vapour line that leans forward (a positive
$ds/dT$ along saturation), so an expansion from saturated vapour ends to the right
of the dome, dry, with no superheat needed, and the cycle can even benefit from a
recuperator that uses the still-warm turbine exhaust to preheat the pumped liquid.
An \engterm{isentropic} fluid sits at the boundary, its vapour line nearly
vertical, so expansion tracks the dome and ends near saturation. The molecular
reading is that heavier, more complex molecules have more internal modes, a larger
specific heat, and so a vapour line that leans forward; light molecules like water
lean back. The organic Rankine cycle exists because the working-fluid menu can be
searched for a dry or isentropic fluid whose critical temperature and saturation
pressures match a chosen source and sink, a freedom the steam cycle, locked to one
substance, never has. The cost of that freedom is that the menu is constrained by
flammability, toxicity, chemical stability at the evaporator temperature, and the
global-warming potential the Kigali schedule now bounds, so the fluid that
maximises efficiency is rarely the fluid a real plant installs.
\end{mastery}

\section{Refrigeration cycles}\pathtag{standard}

A refrigerator moves heat from a cold reservoir to a hot one. The
working cycle is the inverse of a heat-engine cycle; the working
configurations are vapour-compression (the dominant industrial and
domestic refrigerator), absorption (used where waste heat is
available cheaper than electricity), and gas-cycle (used in
specialised applications including aerospace and cryogenics).

\engterm{Vapour-compression cycle}. The cycle consists of four
processes (matching the worked example in chapter~2):

\begin{enumerate}[leftmargin=*]
  \item Compressor (1 to 2): the refrigerant vapour at evaporator
    pressure is compressed to condenser pressure. Specific work
    input $w_c = h_2 - h_1$.
  \item Condenser (2 to 3): the high-pressure vapour rejects heat to
    the high-temperature reservoir and condenses to saturated liquid.
    Specific heat rejected $q_h = h_2 - h_3$.
  \item Expansion valve (3 to 4): the saturated liquid is throttled
    to evaporator pressure, partly flashing to vapour. The throttle
    is isenthalpic: $h_4 = h_3$.
  \item Evaporator (4 to 1): the two-phase mixture absorbs heat from
    the low-temperature reservoir and boils to saturated vapour.
    Specific heat absorbed $q_c = h_1 - h_4$.
\end{enumerate}

The COP for cooling is $\text{COP}_R = q_c/w_c$; for heating (heat
pump), $\text{COP}_{HP} = q_h/w_c$.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-vapor-compression-schematic.tex}

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-vapor-compression-ph.tex}

Figure \ref{fig:vol04ch03:vapor-compression-schematic} shows the four
components as a physical loop, and figure
\ref{fig:vol04ch03:vapor-compression-ph} plots the cycle on the
pressure-enthalpy plane, where the throttle reads as a vertical line and
the evaporator and condenser read as horizontal segments. The companion
\texttt{vapor\_compression.py} takes the four state enthalpies from the
\texttt{r134a\_saturation.csv} table and returns the cooling and heating
COPs.

The choice of refrigerant is set by the temperature range and the
environmental and safety constraints. Historical refrigerants
include ammonia (NH$_3$, high COP, toxic, used in industrial
refrigeration), CFCs and HCFCs (R-12, R-22, banned due to ozone
depletion), HFCs (R-134a, R-410A, high global-warming potential,
being phased out in many jurisdictions), HFOs (R-1234yf, low GWP,
the current automotive air-conditioning standard), natural
refrigerants (CO$_2$, isobutane, propane, ammonia, water for very-
high-temperature applications). The Kigali Amendment
\cite{law:kigali-amendment-2016} to the Montreal Protocol
\cite{law:montreal-protocol-1987} sets a global schedule for phasing
down HFCs over the coming decades. The \texttt{refrigerants.csv} table
collects the global-warming potential, ozone-depletion potential, and
ASHRAE safety class of the common working fluids, current as of 2024;
the World Health Organization performance specification for
vaccine-storage refrigerators \cite{web:who-pqs-refrigerators-2024}
illustrates how a single application fixes the refrigerant and the
operating envelope together.

Selecting a refrigerant is a search across several properties that pull against
one another, and naming them shows why no single fluid is best and why the field
keeps changing. The \engterm{latent heat} of vaporisation sets the cooling effect
per unit mass, so a high latent heat means a smaller mass flow for a given duty; a
fluid like ammonia, with an exceptionally high latent heat, moves a great deal of
cold per kilogram, which is why industrial plants tolerate its toxicity. The
\engterm{saturation pressure} at the evaporator and condenser temperatures sets
the operating pressures and the compressor pressure ratio through the
Clausius-Clapeyron relation of the closing mastery box; a fluid whose saturation
pressures at the working temperatures are neither so low that air leaks inward nor
so high that the vessels grow heavy is preferred, which narrows the menu for each
temperature range. The \engterm{vapour density} at the compressor inlet sets the
volume the compressor must swallow, so a denser vapour allows a smaller
compressor, a lever that favours the higher-pressure fluids. The critical
temperature must sit well above the condenser temperature for a subcritical cycle,
which is what pushed CO$_2$, with its low $31\,\degC$ critical temperature, into
the transcritical operation of an earlier worked example. Layered on top of these
thermodynamic properties are the constraints that have driven three generations of
refrigerant turnover: the ozone-depletion potential that retired the CFCs and
HCFCs, the global-warming potential that the Kigali schedule is now retiring the
HFCs for, the flammability that complicates the hydrocarbons safe in a sealed
domestic charge but hazardous in a large occupied building, and the toxicity that
confines ammonia to machinery rooms. The fluid a plant installs is the one that
satisfies the safety and environmental constraints while coming closest to the
thermodynamic ideal, which is rarely the thermodynamic optimum and never the same
fluid for two different duties.

\begin{mastery}
The pinch temperature that separates the working fluid from each reservoir is set
by the heat exchanger's size, and the effectiveness-NTU relation says how sharply
COP is bought by area. A heat exchanger's \engterm{number of transfer units} is
$\text{NTU} = UA/\dot C_{\min}$, the conductance $UA$ measured against the smaller
stream's heat-capacity rate, and for an evaporator or condenser boiling or
condensing at constant temperature the effectiveness rises as
$\varepsilon = 1 - e^{-\text{NTU}}$. The approach temperature, the pinch between
the refrigerant and the reservoir, is $(1 - \varepsilon)$ times the inlet
difference, so it shrinks exponentially with area: the first units of $\text{NTU}$
buy a large reduction in pinch, and past $\text{NTU} \approx 3$, where
$\varepsilon = 0.95$, further area buys almost nothing. Each kelvin of pinch
removed lets the evaporator run a kelvin warmer or the condenser a kelvin cooler,
which narrows the compressor's temperature lift and raises the COP, by the Carnot
sensitivity $d(\text{COP})/\text{COP} \approx d(\text{lift})/\text{lift}$ that the
opening section derived. The design consequence is a diminishing return: doubling
a small heat exchanger's area lifts the COP substantially, while doubling an
already generous one lifts it negligibly, so the economic optimum sizes each
exchanger to the knee of the effectiveness curve, near $\text{NTU} \approx 2$ to
$3$, where the pinch is a few kelvin and the marginal area no longer pays. This is
why the domestic refrigerator of the teardown, built down to a price with cramped
coils, runs large pinches and a poor COP, while an industrial plant, where energy
cost justifies the metal, sizes its exchangers generously and runs pinches of a
kelvin or two.
\end{mastery}

\engterm{Absorption refrigeration} uses a thermal driving force in
place of mechanical compressor work. The basic ammonia-water
absorption cycle: ammonia vapour evaporates at low pressure (the
cooling effect), is absorbed into water at low temperature, the
ammonia-water solution is pumped to high pressure, the ammonia is
boiled out of the solution by heating, the ammonia vapour condenses
at high pressure (the heat rejection), and the cycle closes through
a throttle. The work input is the small pumping work plus the
thermal energy that drives the ammonia out of solution. Absorption
chillers driven by waste heat or solar heat are economical where the
heat is available at low cost; their COPs are lower than vapour-
compression but their primary-energy efficiency can be higher when
the heat input would otherwise be discarded.

The absorber and generator are together a \engterm{thermal compressor}, and
seeing them that way explains where the cycle's work went and why it needs three
temperatures instead of two. A mechanical compressor raises a vapour's pressure by
doing shaft work on it; the absorption pair does the same job by dissolving the
low-pressure vapour into a liquid, pumping the liquid (cheap, because a liquid is
nearly incompressible, which is the whole reason the scheme pays), and then boiling
the vapour back out at high pressure. The pump work is tiny, but the boiling
demands a heat input at the generator temperature $T_{\text{gen}}$, and that heat
is the true cost. The cycle therefore straddles three temperatures rather than the
two a vapour-compression machine works between: the evaporator at $T_{\text{evap}}$
where cooling is produced, the ambient $T_0$ where the condenser and absorber
reject heat, and the elevated $T_{\text{gen}}$ where the driving heat enters. The
reversible bound is the product form the absorption worked example used, a Carnot
engine extracting work from the driving heat between $T_{\text{gen}}$ and $T_0$
feeding a Carnot refrigerator lifting from $T_{\text{evap}}$ to $T_0$, and it makes
two design constraints visible. The driving heat must be hot enough to boil the
refrigerant out of the strong solution, which sets a floor on $T_{\text{gen}}$
below which the machine will not run at all, and the richer the solution and the
deeper the cold, the higher that floor climbs. A single-effect machine runs on
$85$ to $95\,\degC$ hot water and reaches a thermal COP near $0.7$; a double-effect
machine adds a second generator stage driven by higher-temperature steam and
reuses the heat once internally, reaching $1.1$ to $1.4$, the same
staged-heat-reuse idea the reheat and regenerative power cycles turned on. The
lithium-bromide and water pair serves building air-conditioning, where the
refrigerant is water and the coldest reachable temperature is limited to a few
degrees above freezing; the ammonia and water pair reaches below freezing for
industrial and absorption-refrigerator duty, at the cost of needing a rectifier to
keep water vapour out of the ammonia stream.

\engterm{Gas-cycle refrigeration} (reversed Brayton, Stirling, Joule-
Thomson) operates with a permanent gas (helium, hydrogen, air) and
is the working choice for cryogenic applications below $-100\,\degC$
where conventional refrigerants are unavailable. Aerospace cabin air
conditioning, helium liquefaction (for superconducting magnets and
fusion experiments), and natural-gas liquefaction all use gas-cycle
machinery.

The reversed Brayton cycle makes plain why the gas cycle trades COP for
temperature reach. A refrigerant boiling and condensing exchanges its heat
isothermally, so its cycle hugs the two reservoir temperatures; a permanent gas
has no phase change, so it must absorb and reject heat while its own temperature
slides across a range, and that gliding heat exchange is intrinsically less
efficient than the isothermal exchange of a two-phase fluid. The reversed Brayton
COP for cooling, with all processes ideal and constant specific heats, is
\[
\text{COP}_{\text{rev-B}} = \frac{1}{r_p^{(\gamma-1)/\gamma} - 1},
\]
a function of the pressure ratio alone, exactly the inverted image of the Brayton
engine efficiency $1 - r_p^{-(\gamma-1)/\gamma}$. A modest pressure ratio of
$r_p = 3$ with $\gamma = 1.4$ gives $r_p^{0.286} = 1.369$ and a COP of
$1/(0.369) = 2.7$, already below the vapour-compression cycle's $4$ at a
comparable duty, and the COP falls as the pressure ratio climbs to reach a deeper
cold. The gas cycle accepts that penalty because its virtues answer needs the
vapour-compression cycle cannot: the working gas stays gaseous to arbitrarily low
temperature, where every conventional refrigerant would have frozen solid, so the
cycle can reach the tens of kelvin that liquefy helium; the gas is chemically
inert and non-toxic, which matters in an aircraft cabin where the cooled air is
breathed directly; and a single machine can span a wide temperature range by
staging. Aircraft air-conditioning packs use the reversed Brayton cycle for
exactly the first two reasons, bleeding compressed air from the engine, cooling
it, and expanding it through a turbine whose extracted work both cools the air and
drives the cabin fan, so the cold air delivered to the cabin arrives already
depressurised and breathable.

\engterm{The heat pump and its seasonal balance}. A heat pump is the
vapour-compression cycle run for the heat it rejects rather than the cold
it produces, and its design problem is governed by a fact the steady
COP hides: both the building's demand and the machine's ability to meet
it move with the outdoor temperature, and they move in opposite
directions. As it gets colder the building loses heat faster, so the
demand rises; at the same time the compressor must pump heat across a
wider temperature lift, so the delivered capacity and the COP both fall.
Figure \ref{fig:vol04ch03:heat-pump-balance} plots the two trends. They
cross at the \engterm{balance point}, the outdoor temperature below which
the heat pump can no longer carry the load alone and a supplemental
source, usually electric resistance heat at $\text{COP}=1$, must make up
the shortfall. Every hour spent below the balance point dilutes the
seasonal average performance with resistance heat, which is why the
single steady COP overstates what a real installation delivers over a
winter.

The number that captures the whole season is the \engterm{seasonal
coefficient of performance}, the total heat delivered over a heating
season divided by the total electricity drawn, supplemental heat
included. A unit with a steady COP of $4$ at mild conditions may post a
seasonal COP near $2.5$ once cold hours, defrost cycles, and resistance
backup are summed, the gap depending on the climate and on where the
balance point was set. Setting the balance point is the central sizing
decision: a larger heat pump pushes the balance point colder and leans
less on resistance heat, at higher capital cost and a penalty in mild
weather where it cycles on and off; a smaller unit is cheaper and runs
efficiently in the shoulder seasons but surrenders more cold hours to
expensive backup. Cold-climate heat pumps with variable-speed
compressors and improved low-temperature capacity push the balance point
toward $-15\,\degC$ or below, which is what makes the technology viable
in heating-dominated climates that resistance-backed single-speed units
could not serve economically. The field-log sample in
\texttt{heat\_pump\_field\_log.csv} records the capacity and COP fade with
outdoor temperature that the figure idealises.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-heat-pump-balance.tex}

The balance point sets the technical limit; the economic case turns on the
price of what the heat pump displaces. A heat pump's delivered heat costs the
electricity price divided by the field COP, so against pure electric resistance
heat, which costs the full electricity price for every delivered joule, the
heat pump wins by a factor near the COP in mild weather and by less as the COP
falls (figure \ref{fig:vol04ch03:heat-pump-seasonal-cost}). The seasonal saving
is the area between the two cost curves, weighted by the hours the climate
spends in each temperature bin. A mild maritime climate, where most heating
hours fall above the balance point and the displaced fuel is expensive
electricity, rewards the heat pump heavily; a cold continental climate with
cheap gas displaces a cheaper fuel and spends more hours below the balance
point, narrowing the case. The comparison against gas is the one most
customers actually face, and there the heat pump's advantage rests not on cost
per joule alone but on the carbon intensity of the grid, which the closing
judgment exercise of the chapter takes up.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-heat-pump-seasonal-cost.tex}

\begin{mastery}
The pressure ratio a refrigerant cycle must work across is not a free choice;
it is fixed by the saturation curve of the working fluid between the evaporator
and condenser temperatures, and the Clausius-Clapeyron relation says how
steeply that ratio climbs with the temperature lift. Along the
liquid-vapour saturation line,
\[
\frac{d(\ln p)}{dT} = \frac{h_{fg}}{R T^2},
\]
so integrating between the evaporator temperature $T_c$ and the condenser
temperature $T_h$, with the latent heat $h_{fg}$ and the refrigerant gas
constant $R$ taken roughly constant, gives a pressure ratio
\[
\frac{p_h}{p_c} \approx \exp\!\left[\frac{h_{fg}}{R}
\left(\frac{1}{T_c} - \frac{1}{T_h}\right)\right].
\]
The exponential is the reason a deep lift is punishing: a domestic
refrigerator lifting R-134a from $-10\,\degC$ to $40\,\degC$ works across a
pressure ratio near $4$, while a freezer reaching $-30\,\degC$ against the same
condenser faces a ratio near $12$, a tripling that the worked two-stage example
addressed by splitting the compression. A refrigerant with a high latent heat
relative to its gas constant, the ratio $h_{fg}/R$ in the exponent, climbs in
pressure ratio faster for a given lift, which is one of the property trade-offs
behind selecting a working fluid: ammonia's large $h_{fg}$ gives it an
excellent cooling effect per unit mass but a steep pressure-ratio penalty at
deep lift, while a fluid chosen for a gentle saturation slope eases the
compressor at the cost of a larger volume flow. The saturation curve, read
through Clausius-Clapeyron, is the single property that ties the refrigerant
choice, the compressor pressure ratio, and the staging decision together.
\end{mastery}

\subsection{A domestic refrigerator, taken apart}\pathtag{standard}

The three end-to-end case studies of this chapter sized large machines; the
refrigerator humming in a kitchen is the cheapest cycle most readers own, and
opening its energy budget shows how far a real vapour-compression machine sits
from the bound and where the gap lives. Take a $250\,\text{L}$ combined
fridge-freezer rated at $300\,\text{kWh}/\text{year}$, which is $0.82\,\text{
kWh}/\text{day}$, or an average electrical draw of $34\,\text{W}$ across the
day once the on-off duty cycle is averaged. The cabinet holds its fresh-food
compartment near $4\,\degC$ and its freezer near $-18\,\degC$ in a $22\,\degC$
kitchen.

The first question is what heat the machine must remove, because that load,
not the compressor, sets the daily energy. Heat leaks in through the insulated
walls, around the door gasket, on every door opening, and from the internal
defrost heater and lamp. For a cabinet of this size with a few centimetres of
polyurethane foam, the steady wall conduction runs near $25\,\text{W}$, door
openings and gasket infiltration add perhaps $10\,\text{W}$ averaged over a
day of normal use, and the defrost and internal loads a few watts more, for a
total heat load near $40\,\text{W}$ that the evaporator must pull out. The
upper bar of figure \ref{fig:vol04ch03:fridge-teardown} splits that load, and
its message is that the wall dominates: the single most effective change a
designer can make is thicker or better insulation, because it shrinks the load
the whole machine exists to answer.

The second question is how hard the compressor must work to move that load,
and here the Carnot bound sets the floor. The freezer compartment fixes the
evaporator temperature: to pull heat out of a $-18\,\degC$ space the
refrigerant must boil below it, say at $-25\,\degC$ ($248\,\si{\kelvin}$),
and to reject heat to a $22\,\degC$ kitchen the condenser must run above it,
say at $40\,\degC$ ($313\,\si{\kelvin}$). The Carnot cooling COP for that
device span is $248/(313 - 248) = 3.8$, so the reversible minimum work to move
a $40\,\text{W}$ load is $40/3.8 = 10.5\,\text{W}$. The machine actually draws
$34\,\text{W}$, so its realised COP is $40/34 = 1.2$, a second-law efficiency
near $1.2/3.8 = 0.31$. The lower bar of figure
\ref{fig:vol04ch03:fridge-teardown} splits the $34\,\text{W}$ input: about a
fifth is the reversible minimum, while the small hermetic compressor and its
motor, the throttle, and the finite-temperature pinch in the cramped condenser
and evaporator each claim a comparable share. A domestic refrigerator is a
poor exergy machine because every component is built down to a price, the
heat exchangers are small enough to fit behind the cabinet, and the compressor
is a sealed unit chosen for cost and silence rather than efficiency.

The teardown points the retrofit. The largest single lever is the insulation,
because it cuts the $40\,\text{W}$ load that fixes everything downstream; the
second is the condenser, because enlarging it narrows the pinch and lowers the
condensing temperature, which raises the device COP toward its reservoir
bound. Replacing the throttle with an expander, the move that pays on a large
industrial plant, recovers only a watt or two here and never earns its
mechanical cost at domestic scale. Raising the freezer set-point a few degrees,
finally, lifts the evaporator temperature and raises the Carnot bound itself,
the cheapest gain of all where the stored food tolerates it. The same
second-law accounting that ranked the components of a $400\,\text{MW}$ plant
ranks the parts of a kitchen appliance; only the scale and the price of each
fix differ.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-fridge-teardown.tex}

\begin{mastery}
The on-off cycling that the daily-average draw hides carries its own penalty,
and a short accounting shows why a variable-speed compressor earns its extra
cost. A single-speed refrigerator meets a $40\,\text{W}$ load with a
compressor sized for the worst case, perhaps $120\,\text{W}$ of cooling, so it
runs a third of the time and is off the rest. Each restart pulls the
evaporator down from a warmed state and runs the compressor briefly at an
unfavourable pressure ratio before the cycle settles, and each shutdown leaves
the high and low sides to equalise through the throttle, discarding the
pressure difference the last seconds of running built. These start-stop losses
add a few percent to the energy a steadily running machine would use for the
same average load, and they grow with the cycling frequency. A variable-speed
compressor instead runs continuously at the low speed that exactly matches the
$40\,\text{W}$ load, which raises the COP on two counts: it works at a lower
average pressure ratio because the evaporator stays warmer and steadier, and
it avoids the restart and pressure-equalisation losses entirely. The same
logic governs the cold-climate heat pumps of the previous section, where
variable-speed operation both holds capacity at low outdoor temperature and
eliminates the defrost-driven cycling that drags the seasonal COP down. The
control strategy, not just the cycle thermodynamics, sets what a real machine
delivers over a duty that is never steady.
\end{mastery}

\subsection{A vignette: the thermoacoustic engine and refrigerator}\pathtag{mastery}

Every cycle in this chapter so far has carried its working fluid around a loop of
pipes and vessels past pistons, turbines, or valves. A thermoacoustic machine
dispenses with all of it: the working fluid oscillates in place inside a resonant
tube, and a sound wave does the work a piston would. The device is a closed tube
holding a gas at high mean pressure, with a porous \engterm{stack} of plates set
between a hot and a cold heat exchanger partway along it. When the temperature
gradient across the stack is steep enough, a standing acoustic wave grows
spontaneously, and the machine runs as an engine, converting the heat that
maintains the gradient into acoustic power. Drive the same tube with a
loudspeaker instead, and the wave pumps heat up the stack against the gradient:
the machine runs as a refrigerator.

The thermodynamics is the Stirling cycle carried out by a parcel of gas that
never leaves the stack. As the standing wave oscillates, a gas parcel in a stack
pore moves back and forth along the plate, and its pressure rises and falls a
quarter-cycle out of step with its motion. On the half-cycle when the parcel is
compressed it is also displaced toward the hot end, so it deposits heat into the
plate there; on the return half-cycle it expands and is displaced toward the cold
end, absorbing heat from the plate. Each parcel thus shuttles a small quantity of
heat a short distance along the stack, from cold plate to hot plate, and the
bucket-brigade of parcels down the stack carries heat the whole length. The
spacing of the stack plates is the design crux: the plates must sit a few thermal
penetration depths apart, close enough that a parcel exchanges heat with the plate
within an acoustic cycle but not so close that viscous drag on the oscillating gas
wastes the acoustic power. That single length scale, the thermal penetration
depth $\delta_\kappa = \sqrt{2\alpha/\omega}$ set by the gas thermal diffusivity
$\alpha$ and the acoustic angular frequency $\omega$, governs the whole machine,
and it is the reason thermoacoustic devices run at high mean pressure and in
light gases like helium, where the penetration depth and the acoustic power
density are favourable.

The performance sits below the cycle machines this chapter has ranked. A
thermoacoustic engine reaches a second-law efficiency near $0.2$ to $0.4$, and a
thermoacoustic refrigerator a similar fraction of its Carnot COP, because the
irreversible heat transfer across the finite gap between gas and plate, and the
viscous dissipation in the oscillating boundary layer, both scale with the same
acoustic amplitude that carries the useful power. The machine cannot narrow one
without narrowing the other, the recurring tension between rate and reversibility
that the chapter's opening archetype named. What buys the thermoacoustic machine
its niche is the absence of any sliding seal or lubricated bearing in the cold
part: the only moving part can be a single loudspeaker cone or, in a
travelling-wave design, none at all. That reliability has carried thermoacoustic
coolers into spacecraft instrument cooling and into liquefied-natural-gas
research plants, and thermoacoustic engines into radioisotope-powered generators
where a mechanism that runs for decades without maintenance outranks the modest
efficiency. The device closes the chapter's survey where it began, at the Carnot
bound, reached here not by a turbine or a piston but by a standing wave in a tube
of hot gas.

\subsection{A vignette: ocean thermal energy conversion}\pathtag{mastery}

The cycles of this chapter chase a large temperature span, because the Carnot
factor $1 - T_c/T_h$ rewards it. Ocean thermal energy conversion takes the
opposite bet: it runs a power cycle across the smallest span any serious engine
works with, the twenty-odd kelvin between warm tropical surface water and the cold
water a kilometre below, and makes its case on the sheer size and steadiness of
the resource rather than on efficiency (figure
\ref{fig:vol04ch03:otec-schematic}).

The closed-cycle machine is an ammonia Rankine loop. Warm surface seawater near
$26\,\degC$ boils ammonia in an evaporator; the vapour drives a turbine; cold
water pumped from depth near $5\,\degC$ condenses the ammonia in a condenser; a
feed pump closes the loop. The Carnot bound between $299\,\si{\kelvin}$ and
$278\,\si{\kelvin}$ is $1 - 278/299 = 0.070$, seven percent, and a real cycle
across such a narrow span, with finite pinches in the two heat exchangers eating
into the already small lift, reaches a gross cycle efficiency near four to five
percent. From that must be subtracted the pumping power for the two seawater
flows, which is the number that governs the whole design: moving enough cold water
up a kilometre-long pipe to condense the ammonia can consume a third to a half of
the gross output, so the net efficiency lands near three percent. That
pumping-dominated economy is why every OTEC design lives or dies on the diameter
and depth of its cold-water pipe, a structure ten metres across and a kilometre
long, and why the plants that have run have been demonstration-scale.

Three percent sounds fatal until the resource is weighed. The warm surface layer
of the tropical oceans absorbs solar energy across an area measured in tens of
millions of square kilometres, and it holds its temperature day and night and
season to season, so an OTEC plant delivers baseload power where a solar panel
delivers intermittent power. The economics improve further when the cold deep
water, once it has condensed the ammonia, is sold a second time: it chills
buildings directly as district cooling, and its nutrient load feeds mariculture.
A plant read as a power station alone cannot justify its cold-water pipe; read as
a combined power, cooling, and aquaculture facility drawing on a free and endless
temperature difference, it can. OTEC is the chapter's clearest case that cycle
efficiency is not the figure of merit a system is judged by; the delivered value
per unit of capital against a free resource is.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-otec-schematic.tex}

\subsection{A vignette: magnetocaloric refrigeration}\pathtag{mastery}

Every refrigerator in this chapter has moved heat by compressing and expanding a
fluid. A magnetocaloric machine moves heat by magnetising and demagnetising a
solid, and it carries no volatile refrigerant at all, which is what draws
attention to it as the vapour-compression cycle's environmental bill comes due.

The effect rests on the entropy of the magnetic spins in a suitable solid. In a
material near its magnetic ordering temperature, applying a strong magnetic field
aligns the spins and removes magnetic entropy; if the field is applied fast enough
that no heat escapes, the total entropy must hold constant, so the entropy the
spins lost reappears as lattice-vibration entropy and the solid heats up (figure
\ref{fig:vol04ch03:magnetocaloric-st}). Remove the field adiabatically and the
reverse happens: the spins disorder, drawing entropy back from the lattice, and
the solid cools. The cycle is then the magnetic analog of the gas refrigeration
cycles, with the applied field playing the part of pressure. The solid is
magnetised adiabatically and warms; it rejects that heat to the warm side while
the field is held; it is demagnetised adiabatically and cools; and it absorbs heat
from the cold load while the field is off. A regenerator, usually the working
solid itself arranged as a porous bed with a fluid oscillating through it, carries
heat along the bed so that the small per-cycle temperature swing of a few kelvin
is cascaded into a useful span of twenty or thirty kelvin, exactly the
regenerator role the Stirling and Ericsson cycles turned on.

The appeal is threefold. The working substance is a solid with no ozone-depletion
or global-warming liability, sidestepping the Kigali schedule that is retiring the
fluorinated refrigerants; the process is intrinsically reversible in the limit of
slow field change and good heat transfer, so a magnetic refrigerator can in
principle reach a higher fraction of Carnot than a throttling vapour cycle; and it
runs quietly with few moving parts. The obstacles are equally concrete. The
strong field needs a permanent-magnet assembly or a superconducting coil, both
costly; the best room-temperature materials undergo their large entropy change
only across a narrow temperature window, so a practical bed layers several
materials with staggered ordering temperatures; and moving heat in and out of a
solid fast enough for a useful power density strains the same heat-transfer surface
that every regenerated cycle depends on. Room-temperature magnetic refrigeration
sits today where thermoacoustic cooling sat a generation ago, a laboratory and
demonstration technology whose case is made by the refrigerant it does not use
rather than by a COP that yet beats the machine it would replace.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-magnetocaloric-st.tex}

\subsection{A vignette: the pulse-tube cryocooler}\pathtag{mastery}

The gas-cycle refrigerators of this chapter all reached low temperature by
expanding a gas through a moving part, a turbine or a displacer, sitting in the
cold region. The pulse-tube cryocooler reaches the same cold with no moving part at
the cold end at all, and that single omission is what has carried it into the
duties where a mechanism that fails in a decade is not tolerable.

The machine is a Stirling-type regenerative cooler with the cold displacer replaced
by a column of gas. Figure \ref{fig:vol04ch03:pulse-tube-schematic} lays out the
train: a pressure oscillator at the warm end, a regenerator, a cold heat exchanger
that takes the load, the pulse tube itself, a hot heat exchanger, and an orifice
opening into a reservoir. The pressure oscillator drives the working gas, almost
always helium, back and forth through the regenerator, which stores and returns the
gas's heat as it passes, exactly the regenerator role the Stirling and Ericsson
cycles turned on. In a Stirling cooler a second piston, the cold displacer, would
set the phase between the gas's pressure and its motion so that the gas arrives at
the cold exchanger while expanding and cool. The pulse tube achieves that phase
shift without a cold piston: the orifice and reservoir at the far warm end throttle
the gas that oscillates into and out of the reservoir, and the pressure drop across
the orifice sets the phase between the pressure swing and the flow in the tube. A
gas parcel in the tube is thereby made to compress while moving toward the hot end,
depositing heat there, and to expand while moving toward the cold end, absorbing
heat there, so the column of gas carries heat up the tube from cold exchanger to hot
exchanger, the work a displacer would have done accomplished by a standing pressure
wave and a well-tuned orifice.

The performance sits below a good Stirling cooler. The orifice that sets the phase
also dissipates the gas's kinetic energy as it throttles, so a pulse tube reaches a
second-law efficiency lower than a machine whose displacer recovers that work, and
the temperature it can reach in one stage is limited by the regenerator's falling
heat capacity toward the low tens of kelvin. Two-stage and three-stage pulse tubes,
with a regenerator matrix graded for the cold it must hold, reach a few kelvin, cold
enough to condense helium and to cool superconducting magnets and infrared
detectors. What buys the pulse tube its place, against that modest efficiency, is
the reliability the missing cold displacer confers: with no rubbing seal or bearing
in the cold region, the only wearing part is the warm pressure oscillator, which can
be a flexure-borne linear motor with a maintenance-free life measured in years, and
the absence of a moving cold mass makes the machine quiet and low in vibration. A
vibrating cold finger blurs a telescope's image and couples noise into a
superconducting sensor; a pulse tube, holding its detector at a few kelvin for the
mission life of a spacecraft with nothing moving where the cold is made, answers
exactly the need that the more efficient machines with cold moving parts cannot.
The device closes the refrigeration survey where the thermoacoustic and
magnetocaloric vignettes did, at a machine whose case rests not on its coefficient
of performance but on a virtue the cycle geometry confers, here a working life that
outlasts every seal a colder-running rival would need.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-pulse-tube-schematic.tex}

\section{Combined cycles, cogeneration, trigeneration}\pathtag{mastery}

A \engterm{combined cycle} uses two cycles in series: the heat
rejected by the top cycle drives the bottom cycle. The standard
configuration in modern power generation is a Brayton cycle on top
(natural-gas-fired gas turbine) and a Rankine cycle on bottom (steam
turbine driven by a heat-recovery steam generator). The gas turbine's
exhaust at $\sim 600\,\degC$ contains enough thermal energy to
generate steam at $500\,\degC$ and $15\,\text{MPa}$ for the Rankine
bottoming cycle. The combined-cycle efficiency, which is the sum of
the two cycles' efficiencies minus their product (in a careful
formulation that accounts for the recovered heat), reaches $60\,\%$
to $64\,\%$ in modern installations
(figure \ref{fig:vol04ch03:combined-cycle-flow}). Combined-cycle plants
are the most efficient large fossil-fuel power producers.

\input{volumes/04-energy/ch03-heat-engines-refrigerators/figures/fig-combined-cycle-flow.tex}

\engterm{Cogeneration} (also called combined heat and power, CHP)
recovers useful heat from the rejection of a power cycle. A gas
turbine driving an electric generator also produces steam from its
exhaust; the steam is used for district heating, industrial process
heat, or absorption cooling. The combined first-law efficiency of a
well-engineered CHP plant exceeds $80\,\%$ (versus $33\,\%$ to $60\,\%$
for the same gas turbine running electricity-only). The second-law
analysis is more subtle: the recovered heat is delivered at a lower
temperature than the original combustion, so its exergy content is
proportionally smaller. CHP is the right choice where the heat load
matches the temperature and timing of the cycle's rejection;
electricity-only operation is the right choice where it does not.

The figure that a plant operator must know before quoting an $80\,\%$ combined
efficiency is the electricity it gave up to reach it, measured by the
\engterm{power-to-heat ratio} and the electricity lost per unit of heat delivered.
A steam plant asked for process heat can raise its condenser pressure so the exhaust
condenses at a useful temperature, or bleed steam from the turbine before it fully
expands; either move stops the steam from expanding to the cold condenser and so
forfeits the work that last stretch of expansion would have made. The back-pressure
worked example put a number on it: sending a kilogram of steam to a $0.5\,\text{MPa}$
process rather than to a $10\,\text{kPa}$ condenser surrendered about
$630\,\text{kJ}/\text{kg}$ of electricity to deliver about $2150\,\text{kJ}/\text{kg}$
of heat, a lost-electricity-to-delivered-heat ratio near $0.3$. That ratio is the
true price of cogeneration heat, and it is why the honest comparison is not the
plant's first-law efficiency against an electricity-only plant's but the fuel the
cogeneration plant saves against the fuel a separate boiler and a separate power
plant would have burned to make the same heat and the same electricity. When the
displaced separate heat would have come from a ninety-percent boiler and the
displaced electricity from a grid at forty percent, the cogeneration plant's saving
is real and large; when the grid it displaces is itself a sixty-percent combined
cycle and the heat could have come from a heat pump at a coefficient of performance
of three, the arithmetic can reverse, and the cogeneration plant burns more primary
fuel than the alternatives it replaced. The power-to-heat ratio, set by where the
steam is bled and by the plant's turbine configuration, is the lever an operator
turns to track the changing relative value of heat and electricity across a day and
a season.

\begin{mastery}
The exergy accounting named combustion the largest single destroyer of work
potential in the combined-cycle plant, and the reason lies in a quantity distinct
from the heating value the fuel is usually rated by. A fuel's \engterm{chemical
exergy} is the maximum work obtainable by bringing it to chemical equilibrium with
the environment, and for common hydrocarbons it is close to but not equal to the
higher heating value: for methane the chemical exergy is about
$51.8\,\text{MJ}/\text{kg}$ against a higher heating value near
$55.5\,\text{MJ}/\text{kg}$, a ratio of $0.93$, because the combustion products
carry away some entropy the environment will not return. That chemical exergy is
the true ceiling on the work a fuel can do, and a fuel cell approaches it directly
by oxidising the fuel electrochemically at low temperature with little entropy
generated. A flame does the opposite: it releases the fuel's chemical energy as
heat at a temperature far below the adiabatic flame temperature the reaction could
in principle reach, and then delivers that heat to a working gas across a further
temperature gap. Each of those steps generates entropy, and the sum is the roughly
one-quarter to one-third of the fuel's exergy that combustion destroys before any
turbine turns. The number is nearly independent of how good the turbine is, which is
the uncomfortable lesson: a combined-cycle plant can reach fifty-nine percent of the
fuel's energy as work yet still throw away a third of the fuel's work potential in
the flame, because the first law credits the combustion heat in full while the
second law charges it for the temperature gap it crossed. This is why a fuel cell,
which sidesteps the flame, carries a higher exergy ceiling than any combustion
engine, and why the long-run pressure on the combustor is to burn at higher
temperature, closer to the flame's potential, or to leave the flame behind
altogether.
\end{mastery}

\begin{mastery}
The first-law efficiency of cogeneration counts a joule of $90\,\degC$
hot water as equal to a joule of electricity, which a second-law
accounting refuses to do. The \engterm{exergy} of a heat flow $Q$
delivered at temperature $T$, relative to a dead-state environment at
$T_0$, is $Q(1 - T_0/T)$, the maximum work that flow could in principle
produce. A joule of electricity is pure exergy; a joule of $90\,\degC$
($363\,\si{\kelvin}$) heat against a $15\,\degC$ ($288\,\si{\kelvin}$)
environment carries only $1 - 288/363 = 0.21$ joule of exergy. A CHP
plant delivering $40\,\%$ of its fuel as electricity and $45\,\%$ as
$90\,\degC$ heat has a first-law efficiency of $85\,\%$ but a
second-law (exergy) efficiency of roughly $40\,\% + 0.21 \times 45\,\% =
50\,\%$. The first-law figure is the right one when the question is fuel
burned per unit of delivered energy; the exergy figure is the right one
when the question is how much of the fuel's work potential survived.
Quoting the $85\,\%$ against an electricity-only plant's $40\,\%$
compares unlike quantities and flatters the CHP plant.
\end{mastery}

\engterm{Trigeneration} adds refrigeration (typically through an
absorption chiller driven by part of the recovered heat) to the
cogeneration mix, supplying electricity, heating, and cooling from a
single fuel input. Large commercial buildings, hospitals,
universities, and data centres operate trigeneration plants where the
three loads coincide.

The economic logic of trigeneration turns on a mismatch that cogeneration alone
suffers from. A cogeneration plant sells electricity year round but its heat only
in the heating season, so through the summer its recovered heat has no buyer and
is dumped, which is exactly when the plant would otherwise idle its second
product. Trigeneration converts that summer heat into cooling through an
absorption chiller, so the same recovered stream that heats buildings in winter
chills them in summer, and the plant carries a useful second output across the
whole year. The arithmetic is the product of the two stages the absorption worked
example set out: recovered heat at, say, $95\,\degC$ drives a single-effect
chiller at a thermal COP near $0.7$, so a plant recovering a megawatt of exhaust
heat delivers about $700\,\text{kW}$ of cooling from heat that would otherwise be
rejected. A data centre is the sharpest case, because its cooling load runs at
full strength every hour of every season and rises with the same ambient
temperature that would otherwise strand the plant's heat; feeding the waste heat
of the on-site generators into absorption chillers turns a year-round cooling bill
into a use for a year-round waste stream. The design condition for trigeneration
is therefore not that the three loads be large but that they overlap in time: the
plant earns its third output only in the hours when heat is being recovered and
cooling is wanted at once, which is why the buildings that install it are those,
like hospitals and data centres, whose loads coincide rather than alternate.

\engterm{District heating and cooling networks} extend cogeneration
and trigeneration to neighbourhood and city scale. A central plant
supplies hot water (or steam) and chilled water to a network of
consumer buildings through insulated pipes. The economic case
depends on the heat-load density, the climate, and the available fuel
prices; district heating dominates in Scandinavia, eastern Europe,
and parts of East Asia.

The single number that decides whether a network can be built is the
\engterm{linear heat density}, the annual heat sold per metre of pipe trench,
because the trench is the network's dominant capital cost and its dominant heat
loss. A trench carries a roughly fixed cost per metre to dig and to lay insulated
pipe in, and the buried pipe loses heat to the ground at a rate set by its
insulation and the soil, a loss nearly independent of how much heat flows through
it. Both the capital and the loss are therefore charged against whatever heat the
pipe delivers, so a dense city block selling several megawatt-hours per metre per
year carries the trench easily while a spread suburb selling a fraction of that
cannot. The rule of thumb that has governed European network planning is a linear
heat density above roughly $2\,\text{MWh}$ per metre per year for a network to pay,
which is why district heat follows the dense pre-war city and stops at the low
density suburb. The distribution loss itself, the heat that leaks from the pipes
between plant and customer, runs near ten percent in a well-run dense network and
climbs toward a quarter in a sparse one, and it is the reason the industry has moved
from steam to hot water and then to lower supply temperatures with each generation:
a lower supply temperature cuts the pipe loss directly, since the loss scales with
the temperature difference to the ground, and it lets the plant reject its heat at a
lower temperature, which raises the electricity the cogeneration cycle makes
alongside the heat, by the back-pressure trade the worked cogeneration example set
out. The newest networks run supply temperatures low enough to be fed by industrial
waste heat, data-centre rejection, and large heat pumps rather than a combustion
plant at all, which turns the network from a way to distribute one plant's waste
heat into a way to gather many low-grade sources into one warm loop. Chapter~6 of
this volume develops the heat-exchanger and pumping discipline that the network's
design depends on.

\section{Failure: efficiency claims that violated the second law}\pathtag{core}

The history of energy engineering is filled with proposals, products,
and patents that claim cycle efficiencies exceeding the second-law
bound. The proposals are sometimes naive (an inventor who has not
internalised the Carnot bound), sometimes deliberate (a marketing
team selling a low-quality refrigerator with inflated COP claims),
and sometimes the output of a long measurement campaign whose
systematic errors have not been controlled. The chapter's failure
section catalogues the recurring patterns and the diagnostic that
detects each. The companion \texttt{efficiency\_claim\_check.py}
implements the two-step diagnosis for engine, heat-pump, and
combustion-appliance claims, and \texttt{cycle\_efficiency\_benchmarks.csv}
collects the realised efficiencies of the major cycle classes against
their Carnot bounds for the plausibility check of step two.

\subsection{The over-unity heat pump}

A vendor claims a heat pump with COP greater than the Carnot value
for its operating temperatures. The first diagnostic: compute the
Carnot COP for the claimed reservoir temperatures and confirm the
claimed COP exceeds it.

Working numbers: a domestic air-source heat pump operates at $T_h =
35\,\degC$ (condenser interior) and $T_c = -5\,\degC$ (outdoor coil
under typical winter conditions). The Carnot heating COP is $T_h/
(T_h - T_c) = 308/40 = 7.7$. A claimed COP of $4$ to $5$ is plausible
($52$ to $65\,\%$ of Carnot). A claimed COP of $8$ violates the
Carnot bound and is either a measurement error (most often
misidentified reservoir temperatures, or steady-state efficiency
confused with cyclic efficiency including defrost penalties), a
mistake in the COP definition (cooling COP for a heat pump quoted as
heating COP, or the inverse), or an unaccounted input (gas-fired
defrost system, electric supplemental heater).

The diagnostic for a heat-pump claim: identify $T_h$ and $T_c$ from
the actual operating boundary, compute the Carnot COP, and require
the claimed COP to lie below it.

\subsection{The 100\% efficient combustion device}

A vendor sells a heating appliance (a residential furnace, a
commercial boiler) and claims a first-law efficiency exceeding
$100\,\%$. The claim is often supported by the appliance's measured
output divided by the lower heating value (LHV) of the fuel; a
\engterm{condensing} appliance recovers the latent heat of the water
vapour in the combustion products and can deliver more heat than the
LHV figure indicates. The ratio of higher heating value (HHV) to LHV
is approximately $1.11$ for natural gas, so an apparent efficiency of
$108\,\%$ (LHV basis) is $97\,\%$ on an HHV basis and entirely
plausible.

The diagnostic: identify whether the efficiency is quoted on HHV or
LHV basis. An LHV-basis efficiency above the HHV/LHV ratio (a
condensing-furnace efficiency above $111\,\%$, a condensing-boiler
above $110\,\%$) violates the first law; below that ratio, the
appliance is condensing and the claim is consistent. Working
condensing furnaces achieve $90\,\%$ to $97\,\%$ HHV efficiency
($98\,\%$ to $106\,\%$ LHV).

\subsection{The water-fuelled car}

A class of proposals claims to drive an internal-combustion engine on
water by electrolysing the water in situ and burning the resulting
hydrogen-oxygen mixture. The first-law balance: the work required to
electrolyse water is the same as (or, accounting for irreversibilities,
greater than) the heat released by combusting the resulting gases. The
proposed device is a closed loop with no net energy input that
produces work; it is a first-kind perpetual motion machine.

The diagnostic: draw the boundary around the proposed device and ask
where the energy comes from. If the answer is the on-board battery,
the device is a low-efficiency battery-electric vehicle; the
electrolysis-and-recombustion loop is an energy-losing detour. If
the answer is "the water itself," the proposal violates the first
law.

The proposal recurs because hydrogen-oxygen combustion is genuinely
exothermic (relative to the gases), and the electrolysis is
genuinely energy-storing (relative to the gases). The trick is the
direction: combustion releases the energy that electrolysis put in.
A net round-trip cannot produce energy.

\subsection{The Carnot-violating refrigerator}

A consumer product is advertised with a cooling capacity and a power
input that imply a COP higher than the Carnot bound for its
operating temperatures. Working example: a small ``thermoelectric''
refrigerator advertised as cooling $5\,\text{L}$ to $20\,\degC$ below
ambient (so $T_c \approx 0\,\degC$, $T_h \approx 20\,\degC$) at
$3\,\text{W}$ of input. The Carnot COP is $273/20 = 13.6$, so a
$3\,\text{W}$ input could in principle provide $40\,\text{W}$ of
cooling; the advertised capacity might be $30\,\text{W}$. The
advertised numbers are internally consistent with the Carnot bound.
The actual product, of course, achieves perhaps $20\,\%$ of Carnot
(thermoelectrics have poor exergy efficiency), so the real cooling
capacity is closer to $8\,\text{W}$. The error in the marketing is
not a Carnot violation but an unrealistic figure-of-merit assumption.

The diagnostic: separate the Carnot-physics check from the realised-
efficiency check. A claim that violates Carnot is a second-law
violation; a claim that satisfies Carnot but overstates the realised
fraction is a marketing exaggeration or a measurement error.

\begin{principle}[Two-step diagnosis of an efficiency claim]
For any cyclic energy device, the efficiency or COP claim is checked
in two steps. Step one: identify the operating reservoir temperatures
and compute the Carnot bound (efficiency, $\text{COP}_R$, or
$\text{COP}_{HP}$ as appropriate). A claim above the bound violates
the second law; the diagnostic is the bound itself. Step two: if the
claim is below the Carnot bound, identify the realised second-law
efficiency $\eta_{II} = \eta_{\text{claimed}}/\eta_{\text{Carnot}}$
and compare to known performance for the device class. Working
Rankine plants achieve $\eta_{II} \approx 0.6$ to $0.7$; vapour-
compression refrigerators $\approx 0.5$ to $0.7$; internal-combustion
engines $\approx 0.4$ to $0.6$. A claim with implausibly high
$\eta_{II}$ may not violate physics, but it is the diagnostic for a
careful look at the measurement protocol.
\end{principle}

\begin{estimation}
\textbf{Estimate the thermal efficiency of a typical compact-car
spark-ignition engine at its rated operating point.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A compact car gets approximately $7\,\text{L}/100\,\text{km}$ on
gasoline in mixed driving. The energy content of gasoline is roughly
$32\,\text{MJ}/\text{L}$ (lower heating value). The energy delivered
to the fuel-burning engine per kilometre is $7 \times 32/100 = 2.24
\,\text{MJ}/\text{km} = 2240\,\text{kJ}/\text{km}$.

The road power demand at a moderate cruising speed of $100\,\text{km}/
\text{h}$ for a $1500\,\text{kg}$ compact car: rolling resistance at
$C_r \approx 0.01$ gives $F_{\text{roll}} = 0.01 \times 1500 \times
9.81 = 147\,\text{N}$. Aerodynamic drag at $A C_d \approx 0.7\,
\text{m}^2$ and air density $1.2\,\text{kg}/\text{m}^3$ at
$28\,\text{m}/\text{s}$ gives $F_{\text{drag}} = 0.5 \times 1.2 \times
0.7 \times 28^2 = 329\,\text{N}$. Total road force $\sim 480\,\text{N}$;
road power $\sim 480 \times 28 = 13.5\,\text{kW}$.

Per kilometre at $100\,\text{km}/\text{h}$: $13.5\,\text{kW} \times
36\,\text{s} = 487\,\text{kJ}$. The thermal efficiency at the
cruising point is roughly $487/2240 \approx 22\,\%$.

The estimate is a tank-to-wheels number that includes drivetrain
losses ($\sim 15\,\%$ of engine output), so the engine's brake thermal
efficiency at this operating point is closer to $26\,\%$. Including
operation at non-optimal points (city traffic, cold starts, idling),
the lifecycle average drops further.

Modern compact-car gasoline engines have peak brake thermal
efficiencies near $35\,\%$ at the design operating point and average
$25\,\%$ to $30\,\%$ over typical driving cycles. Manufacturer claims
for ``up to $40\,\%$ thermal efficiency'' refer to the design point,
not the average.

The postmortem: the estimate is correctly in the $20\,\%$ to $30\,\%$
range. The largest fragile inputs are the assumed fuel economy
(varies with vehicle and driving) and the aerodynamic drag (depends
on vehicle shape). The thermal efficiency is bounded above by the
Otto-cycle ideal (around $60\,\%$ for $r = 10$), and the real engine
achieves $30\,\%$ to $50\,\%$ of that.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Estimate the thermal efficiency of a household car
engine three ways]
The reader estimates the thermal efficiency of a familiar
spark-ignition or compression-ignition engine through three
independent calculations and compares the three to one another and to
the manufacturer's stated efficiency. The deliverable is a fifteen-
to twenty-five-page report.

\textbf{Track}. Calculation plus data-collection (driving log).
\textbf{Hazard class}: low.

\textbf{Method 1: tank-to-wheels from a driving log}. Over four weeks,
record every refuelling: odometer reading, litres of fuel, fuel
type, route categorisation (city, highway, mixed). Compute average
fuel consumption $\text{L}/100\,\text{km}$. Use the road-load
formula (rolling resistance plus aerodynamic drag) with stated
vehicle parameters to estimate the work delivered to the wheels per
kilometre. Compute the tank-to-wheels efficiency $\eta_1 =
W_{\text{wheels}}/Q_{\text{fuel}}$, and back out the engine's brake
thermal efficiency by accounting for drivetrain and accessory
losses (state the assumed loss fraction).

\textbf{Method 2: rated power and manufacturer fuel-economy
specification}. Read the manufacturer's rated power (kW) and rated
fuel-economy figures from the vehicle handbook. Use the rated test
cycle to back out the engine's brake thermal efficiency at the rated
point.

\textbf{Method 3: air-standard cycle analysis}. From the engine's
compression ratio (printed in the handbook) and the appropriate
air-standard cycle (Otto for spark-ignition, Diesel for compression-
ignition), compute the ideal-cycle thermal efficiency. Apply a
typical brake-to-ideal ratio (state the assumption) to estimate the
actual efficiency.

\textbf{Comparison}. Tabulate the three efficiencies, the
manufacturer's stated efficiency (if available), and discuss the
sources of discrepancy. State which method is most sensitive to
which uncertainty.

\textbf{Reflection ($1{,}500$ to $2{,}500$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
  \item Which method gave the highest estimate; which gave the
    lowest? Why?
  \item Where does the gap from the Carnot bound originate? Identify
    the largest single irreversibility.
  \item One design change that would improve the engine's
    efficiency, and the cost of that change.
\end{itemize}

\textbf{Deliverable}. The driving log, the three calculations, the
comparison table, and the reflection.

\textbf{Time}. Four weeks of passive logging, six to ten hours of
calculation, four to six hours of write-up.

\textbf{Optional extension}. A reader with a household heat pump can
repeat the three-way comparison for the heat pump, logging delivered
heat against electricity and outdoor temperature, in the spirit of the
\texttt{heat\_pump\_field\_log.csv} sample, and compare the field COP to
the Carnot bound at each outdoor temperature.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A Carnot engine operates between $T_h = 600\,\si{\kelvin}$ and $T_c =
300\,\si{\kelvin}$. Find the thermal efficiency. If the heat input is
$1\,\text{MJ}$, find the work output and the heat rejected.

\paragraph{Solution sketch.} The efficiency is $\eta = 1 - T_c/T_h =
1 - 300/600 = 0.5$. The work output is $W = \eta Q_h = 0.5 \times
1\,\text{MJ} = 0.5\,\text{MJ}$. The heat rejected is $Q_c = Q_h - W =
1 - 0.5 = 0.5\,\text{MJ}$. As a check, $Q_c/Q_h = T_c/T_h = 0.5$, the
defining property of the reversible cycle.
\end{exercise}

\begin{exercise}
A steam Rankine cycle has a boiler at $5\,\text{MPa}/450\,\degC$ and
a condenser at $10\,\text{kPa}$. Assuming ideal (isentropic) turbine
and pump, find the thermal efficiency. Use the steam tables.

\paragraph{Solution sketch.} From the steam tables: at the boiler exit
($5\,\text{MPa}$, $450\,\degC$) the superheated vapour has $h_3 \approx
3317\,\text{kJ}/\text{kg}$ and $s_3 \approx 6.82\,\text{kJ}/\text{kg}
\cdot\text{K}$. The turbine is isentropic to $10\,\text{kPa}$, where
$s_f = 0.649$ and $s_{fg} = 7.50$, so the exit quality is $x_4 =
(6.82 - 0.649)/7.50 = 0.82$ and $h_4 = 191.8 + 0.82 \times 2392.8 =
2154\,\text{kJ}/\text{kg}$. The condenser exit is saturated liquid,
$h_1 = 191.8\,\text{kJ}/\text{kg}$, $v_f = 0.00101\,\text{m}^3/
\text{kg}$. The pump work is $w_p = v_f(p_b - p_c) = 0.00101 \times
(5000 - 10) = 5.0\,\text{kJ}/\text{kg}$, so $h_2 = 196.8$. Then $q_h =
h_3 - h_2 = 3120$, $w_t = h_3 - h_4 = 1163$, and $\eta_R = (w_t -
w_p)/q_h = (1163 - 5)/3120 = 0.371$, about $37\,\%$. The exit quality
$0.82$ sits at the low end of acceptable turbine-exit wetness, which is
the practical reason reheat is added.
\end{exercise}

\begin{exercise}
An air-standard Brayton cycle has a pressure ratio of $12$ and a
turbine inlet temperature of $1400\,\si{\kelvin}$. Compressor inlet
is at $300\,\si{\kelvin}$ and $100\,\text{kPa}$. Assuming
$\gamma = 1.4$ and isentropic compression and expansion, find the
thermal efficiency, the net work per unit mass, and the back-work
ratio.

\paragraph{Solution sketch.} With $c_p = 1.005\,\text{kJ}/\text{kg}
\cdot\text{K}$ and exponent $(\gamma-1)/\gamma = 0.286$: $T_2 = 300
\times 12^{0.286} = 610\,\si{\kelvin}$ and $T_4 = 1400 \times
12^{-0.286} = 688\,\si{\kelvin}$. Turbine work $w_t = c_p(T_3 - T_4) =
1.005 \times 712 = 715\,\text{kJ}/\text{kg}$; compressor work $w_c =
c_p(T_2 - T_1) = 1.005 \times 310 = 312\,\text{kJ}/\text{kg}$; net work
$w_{\text{net}} = 403\,\text{kJ}/\text{kg}$. Heat input $q_h = c_p(T_3 -
T_2) = 1.005 \times 790 = 794\,\text{kJ}/\text{kg}$, so $\eta_B =
403/794 = 0.508$, matching $1 - 12^{-0.286}$. The back-work ratio is
$w_c/w_t = 312/715 = 0.44$. The high back-work ratio is the Brayton
cycle's signature. The solver \texttt{brayton\_cycle.py} returns the
same values.
\end{exercise}

\begin{exercise}
An Otto cycle has a compression ratio of $10$, a heat input of
$2000\,\text{kJ}/\text{kg}$, and initial conditions $T_1 = 300\,\si{
\kelvin}$, $p_1 = 100\,\text{kPa}$. With $\gamma = 1.4$, find the
maximum temperature, the maximum pressure, the thermal efficiency,
and the mean effective pressure.

\paragraph{Solution sketch.} Compression: $T_2 = 300 \times 10^{0.4} =
754\,\si{\kelvin}$, $p_2 = 100 \times 10^{1.4} = 2512\,\text{kPa}$.
Constant-volume heat addition with $c_v = 0.718$: $T_3 = T_2 +
q_{\text{in}}/c_v = 754 + 2000/0.718 = 3540\,\si{\kelvin}$ (the maximum
temperature), and $p_3 = p_2 (T_3/T_2) = 2512 \times 3540/754 =
11{,}790\,\text{kPa}$ (the maximum pressure). Efficiency $\eta_O = 1 -
10^{-0.4} = 0.602$, so $w_{\text{net}} = 0.602 \times 2000 = 1204\,
\text{kJ}/\text{kg}$. For the MEP, $v_1 = RT_1/p_1 = 0.287 \times
300/100 = 0.861\,\text{m}^3/\text{kg}$ and $v_2 = v_1/10 = 0.0861$, so
$\text{MEP} = w_{\text{net}}/(v_1 - v_2) = 1204/0.775 = 1554\,\text{kPa}$.
The peak temperature is unphysically high because the air-standard model
ignores dissociation and wall heat loss; \texttt{otto\_cycle.py} returns
these values.
\end{exercise}

\begin{exercise}
A Diesel cycle has a compression ratio of $18$ and a cut-off ratio of
$2.5$. With $\gamma = 1.4$, find the thermal efficiency. Compare to
the Otto-cycle efficiency at the same compression ratio.

\paragraph{Solution sketch.} The Diesel efficiency is
\[
\eta_D = 1 - \frac{1}{r^{\gamma-1}}\cdot\frac{r_c^\gamma - 1}{
\gamma(r_c - 1)}
= 1 - \frac{1}{18^{0.4}}\cdot\frac{2.5^{1.4} - 1}{1.4(2.5 - 1)}.
\]
Here $18^{0.4} = 3.18$, $2.5^{1.4} = 3.61$, so the cut-off bracket is
$(3.61 - 1)/(1.4 \times 1.5) = 2.61/2.1 = 1.24$. Then $\eta_D = 1 -
(1/3.18)(1.24) = 1 - 0.391 = 0.609$, about $61\,\%$. The Otto cycle at
$r = 18$ would give $\eta_O = 1 - 18^{-0.4} = 1 - 0.314 = 0.686$, about
$69\,\%$. The Otto value is higher at equal compression ratio because
its constant-volume heat addition occurs at a higher average
temperature. The Diesel's real advantage is that it can run at $r = 18$
at all, where a gasoline charge would knock.
\end{exercise}

\begin{exercise}
A heat pump operates between $T_c = 270\,\si{\kelvin}$ (outdoor air)
and $T_h = 295\,\si{\kelvin}$ (indoor air). It delivers $10\,\text{kW}$
of heating with an actual COP $50\,\%$ of the Carnot value. Find the
electrical input power and the heat absorbed from the outdoor air.

\paragraph{Solution sketch.} The Carnot heating COP is $T_h/(T_h - T_c)
= 295/25 = 11.8$. The actual COP is half that, $5.9$. The electrical
input is $\dot W = \dot Q_h/\text{COP} = 10\,\text{kW}/5.9 =
1.69\,\text{kW}$. The heat absorbed from the outdoor air is the first
law on the device: $\dot Q_c = \dot Q_h - \dot W = 10 - 1.69 =
8.3\,\text{kW}$. The point of the heat pump is that it delivers
$10\,\text{kW}$ of heating from $1.7\,\text{kW}$ of electricity by
pulling the balance out of cold outdoor air, where a resistive heater
would draw the full $10\,\text{kW}$.
\end{exercise}

\begin{exercise}
A vapour-compression refrigeration cycle uses R-134a with evaporator
at $-10\,\degC$, condenser at $40\,\degC$, isentropic compression,
saturated liquid at condenser exit, saturated vapour at evaporator
exit. Find the COP, the specific cooling capacity, and the
compressor work per kilogram.

\paragraph{Solution sketch.} From \texttt{r134a\_saturation.csv}:
saturated vapour at $-10\,\degC$ has $h_1 = 392.4\,\text{kJ}/\text{kg}$;
the isentropic compression to the $40\,\degC$ condenser pressure reaches
$h_2 \approx 425.6\,\text{kJ}/\text{kg}$; saturated liquid at $40\,\degC$
has $h_3 = 256.4\,\text{kJ}/\text{kg}$, and the throttle gives $h_4 =
h_3 = 256.4$. The specific cooling capacity is $q_c = h_1 - h_4 =
392.4 - 256.4 = 136\,\text{kJ}/\text{kg}$. The compressor work is $w_c =
h_2 - h_1 = 33.2\,\text{kJ}/\text{kg}$. The cooling COP is $q_c/w_c =
136/33.2 = 4.1$. Against the Carnot COP $263/50 = 5.3$, the ideal cycle
reaches $77\,\%$ of Carnot, the loss almost all in the throttle. The
solver \texttt{vapor\_compression.py} returns the same figures.
\end{exercise}

\begin{exercise}
A combined-cycle plant has a gas-turbine top cycle with $\eta_{\text{GT}}
= 38\,\%$ and a steam-turbine bottom cycle that converts $60\,\%$ of
the gas-turbine exhaust heat to additional work, with the bottom
cycle's own thermal efficiency of $35\,\%$. Compute the overall
combined-cycle efficiency.

\paragraph{Solution sketch.} Per unit fuel energy, the gas turbine
produces work $0.38$ and rejects $1 - 0.38 = 0.62$ as exhaust. The
heat-recovery steam generator captures $60\,\%$ of that exhaust,
$0.60 \times 0.62 = 0.372$, as heat into the bottom cycle. The bottom
cycle converts $35\,\%$ of that to work, $0.35 \times 0.372 = 0.130$.
The total work per unit fuel is $0.38 + 0.130 = 0.51$, so the combined
efficiency is $51\,\%$. The equivalent compact form is $\eta_{cc} =
\eta_{\text{GT}} + \eta_{\text{HRSG}}(1 - \eta_{\text{GT}})
\eta_{\text{ST}}$. The result exceeds either cycle alone, which is the
case for the combined cycle.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the Carnot efficiency $\eta = 1 - T_c/T_h$ for a reversible
heat engine between two reservoirs at $T_h$ and $T_c$, using the
entropy balance over the cycle.

\paragraph{Solution sketch.} Over one cycle the working fluid returns to
its initial state, so its net entropy change is zero. For a reversible
engine the only entropy transfers are with the two reservoirs: the fluid
gains $Q_h/T_h$ from the hot reservoir during isothermal heat intake and
loses $Q_c/T_c$ to the cold reservoir during isothermal heat rejection,
with no entropy generated. The cyclic entropy balance is therefore
$Q_h/T_h - Q_c/T_c = 0$, giving $Q_c/Q_h = T_c/T_h$. The first law over
the cycle gives $W = Q_h - Q_c$, so $\eta = W/Q_h = 1 - Q_c/Q_h =
1 - T_c/T_h$. The substitution of the entropy ratio for the heat ratio
is the whole content of the result.
\end{exercise}

\begin{exercise}
Derive the air-standard Brayton-cycle thermal efficiency $\eta = 1 -
r_p^{-(\gamma-1)/\gamma}$ in terms of the pressure ratio $r_p$ and
$\gamma$, treating the compressor and turbine as isentropic.

\paragraph{Solution sketch.} Heat is added at constant pressure $q_h =
c_p(T_3 - T_2)$ and rejected at constant pressure $q_c = c_p(T_4 -
T_1)$, so $\eta = 1 - q_c/q_h = 1 - (T_4 - T_1)/(T_3 - T_2)$. The two
isentropic legs share the same pressure ratio, so $T_2/T_1 = T_3/T_4 =
r_p^{(\gamma-1)/\gamma}$. Write $T_2 = T_1 r_p^{(\gamma-1)/\gamma}$ and
$T_3 = T_4 r_p^{(\gamma-1)/\gamma}$; then $T_3 - T_2 = (T_4 -
T_1)r_p^{(\gamma-1)/\gamma}$. Substituting,
\[
\eta = 1 - \frac{T_4 - T_1}{(T_4 - T_1)r_p^{(\gamma-1)/\gamma}}
= 1 - r_p^{-(\gamma-1)/\gamma}.
\]
The temperatures cancel, leaving the efficiency a function of the
pressure ratio alone.
\end{exercise}

\begin{exercise}
Derive the air-standard Otto-cycle thermal efficiency $\eta = 1 -
r^{-(\gamma-1)}$ in terms of the compression ratio $r$ and $\gamma$.

\paragraph{Solution sketch.} Heat is added and rejected at constant
volume, so $q_h = c_v(T_3 - T_2)$ and $q_c = c_v(T_4 - T_1)$, giving
$\eta = 1 - (T_4 - T_1)/(T_3 - T_2)$. The two isentropic legs share the
same volume ratio $r$, so $T_2/T_1 = T_3/T_4 = r^{\gamma-1}$. Then $T_2 =
T_1 r^{\gamma-1}$ and $T_3 = T_4 r^{\gamma-1}$, so $T_3 - T_2 = (T_4 -
T_1)r^{\gamma-1}$. Substituting,
\[
\eta = 1 - \frac{T_4 - T_1}{(T_4 - T_1)r^{\gamma-1}}
= 1 - r^{-(\gamma-1)}.
\]
The structure is identical to the Brayton derivation, with the volume
ratio replacing the pressure ratio and the exponent $\gamma - 1$
replacing $(\gamma-1)/\gamma$.
\end{exercise}

\begin{exercise}
Show that the Carnot COP for a refrigerator, $\text{COP}_R = T_c/(T_h
- T_c)$, and for a heat pump, $\text{COP}_{HP} = T_h/(T_h - T_c)$,
satisfy $\text{COP}_{HP} = \text{COP}_R + 1$.

\paragraph{Solution sketch.} Subtract directly:
\[
\text{COP}_{HP} - \text{COP}_R = \frac{T_h}{T_h - T_c} -
\frac{T_c}{T_h - T_c} = \frac{T_h - T_c}{T_h - T_c} = 1.
\]
The result also follows from the energy balance: the heat delivered to
the hot side is the heat pulled from the cold side plus the work input,
$Q_h = Q_c + W$. Dividing by $W$ gives $\text{COP}_{HP} = \text{COP}_R +
1$ for any device, reversible or not, so the identity is not special to
the Carnot case.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the maximum thermal efficiency of a coal-fired power plant
with boiler at $560\,\degC$ and cooling-water condenser at $30\,\degC$,
and compare to the typical achieved efficiency ($43\,\%$ for modern
ultra-supercritical plants).

\paragraph{Reference estimate.} Convert to kelvin: $T_h = 833\,\si{
\kelvin}$, $T_c = 303\,\si{\kelvin}$. The Carnot bound is $\eta = 1 -
303/833 = 1 - 0.364 = 0.636$, about $64\,\%$. The achieved $43\,\%$ is
$43/64 = 0.67$ of the Carnot bound, a second-law efficiency near $0.7$,
which is typical for a well-engineered Rankine plant. The gap to Carnot
comes from finite-temperature heat transfer in the boiler and
condenser, irreversibilities in the turbine and pump, and the fact that
heat is added over a range of temperatures rather than isothermally at
$T_h$.
\end{exercise}

\begin{exercise}
Estimate the COP of a typical residential air-source heat pump
operating in a $-10\,\degC$ outdoor environment to heat a $20\,\degC$
indoor space. State the second-law efficiency assumed.

\paragraph{Reference estimate.} In kelvin, $T_h = 293$, $T_c = 263$,
lift $30\,\si{\kelvin}$. The Carnot heating COP is $293/30 = 9.8$. Real
air-source heat pumps run at $40$ to $50\,\%$ of Carnot at this lift
once defrost cycling and the temperature pinch in the heat exchangers
are counted. Taking $\eta_{II} = 0.45$ gives a field COP of $0.45 \times
9.8 = 4.4$. The field-log data in \texttt{heat\_pump\_field\_log.csv}
shows a field COP near $2.2$ at a $-8\,\degC$ outdoor temperature, lower
than this clean estimate because the logged days include defrost
penalties and supplemental resistive heating, which the simple
second-law fraction does not capture. A defensible answer is COP $4$ to
$4.5$ for the steady operating point, falling toward $2$ to $3$ over a
real cold-weather duty cycle.
\end{exercise}

\begin{exercise}
Estimate the thermal efficiency of a household electric oven
considering only the heating of the food (not the oven cavity or
losses to the kitchen).

\paragraph{Reference estimate.} Heating a $1\,\text{kg}$ dish from
$20\,\degC$ to $180\,\degC$, with a food specific heat near $3\,\text{kJ}
/\text{kg}\cdot\text{K}$, needs $3 \times 1 \times 160 = 480\,\text{kJ}$
delivered to the food. A typical oven draws $2\,\text{kW}$ and takes
perhaps $40\,\text{min} = 2400\,\text{s}$ for the bake including
preheat, so it consumes $2 \times 2400 = 4800\,\text{kJ}$. The
food-only efficiency is $480/4800 \approx 10\,\%$. The figure is low
because most of the electricity heats the oven cavity, the door, and
the kitchen, and is lost on every air exchange. This is a first-law
efficiency; there is no Carnot bound to bump against, since resistive
heating converts electricity to heat at $100\,\%$ and the loss is
entirely in where the heat goes, not in the conversion.
\end{exercise}

\begin{exercise}
Estimate the cooling capacity required to maintain a $300\,\text{m}^2$
office space in a $35\,\degC$ climate at indoor $22\,\degC$, and the
corresponding electrical input for an air-conditioner with
$\text{COP} = 3$.
\paragraph{Reference estimate.} A common rule of thumb for an office in
a hot climate is a cooling load near $100$ to $150\,\text{W}/\text{m}^2$
once envelope gain, occupants, lighting, and equipment are summed. Take
$120\,\text{W}/\text{m}^2$: the cooling load is $300 \times 120 =
36\,\text{kW}$ (about $10$ tons of refrigeration). At a seasonal COP of
$3$, the electrical input is $\dot Q_c/\text{COP} = 36/3 = 12\,\text{kW}$.
A sanity check from first principles: the dominant gains are solar
through glazing and internal loads, both of order tens of watts per
square metre, so $36\,\text{kW}$ for a $300\,\text{m}^2$ floor is the
right order. Raising the COP to $4$ with a better unit would cut the
electrical draw to $9\,\text{kW}$.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a Rankine-cycle solver: given boiler pressure, boiler
temperature, condenser pressure, turbine and pump isentropic
efficiencies, return the cycle's thermal efficiency, the specific
work, and the cycle state points. Verify against textbook values.

\paragraph{Model-answer sketch.} Structure the solver as four steps
matching the four components. Pump: with feedwater enthalpy and specific
volume from the saturated-liquid line at $p_c$, compute $h_2 = h_1 +
v_f(p_b - p_c)/\eta_p$. Boiler: read $h_3$, $s_3$ from the superheated
table at $(p_b, T_b)$. Turbine: find the isentropic exit state at $s_4 =
s_3$ and $p_c$ (compute the quality if inside the dome), get $h_{4s}$,
then apply $h_4 = h_3 - \eta_t(h_3 - h_{4s})$. Condenser: $h_1$ from the
saturated-liquid line at $p_c$. Then $\eta = (h_3 - h_4 - (h_2 -
h_1))/(h_3 - h_2)$. Verify on exercise 2: the ideal case
($\eta_t = \eta_p = 1$) should return $\eta \approx 0.37$ and exit
quality $\approx 0.82$. A non-isentropic turbine at $\eta_t = 0.85$
drops the efficiency to roughly $0.32$ and raises the exit quality,
because the irreversible expansion ends further inside the superheated
or higher-quality region.
\end{exercise}

\begin{exercise}
Implement an Otto-cycle simulator using a finite-rate combustion
model (Wiebe function) instead of instantaneous heat addition.
Compare the simulated thermal efficiency to the air-standard ideal
for the same compression ratio and heat input.

\paragraph{Model-answer sketch.} Replace the instantaneous
constant-volume heat addition with a heat-release schedule following the
Wiebe function, the cumulative burned-mass fraction
\[
x_b(\theta) = 1 - \exp\!\left[-a\left(\frac{\theta -
\theta_0}{\Delta\theta}\right)^{m+1}\right],
\]
with typical $a = 5$, $m = 2$, burn duration $\Delta\theta \approx
40$ to $60$ crank-angle degrees. Integrate the first law over crank
angle with the cylinder volume from the slider-crank kinematics, adding
$dQ = q_{\text{in}} \, dx_b$ at each step and updating temperature and
pressure for the instantaneous volume. The finite burn duration spreads
heat release across the early expansion rather than concentrating it at
top dead centre, so the simulated thermal efficiency falls a few points
below the air-standard $1 - r^{-(\gamma-1)}$, the gap widening as the
burn duration lengthens. Adding a wall-heat-loss term widens it further.
The exercise's lesson is that the air-standard ideal is an upper bound
that real combustion timing erodes.
\end{exercise}

\begin{exercise}
Implement a vapour-compression cycle simulator using a refrigerant
property-table look-up (or CoolProp). Given $T_{\text{evap}}$,
$T_{\text{cond}}$, isentropic compressor efficiency, sub-cooling at
condenser exit, and superheat at evaporator exit, return the cycle
COP and the four state points. Verify on the worked R-134a example.

\paragraph{Model-answer sketch.} State 1 is at the evaporator pressure
with the specified superheat above $T_{\text{evap}}$; read $h_1$, $s_1$.
For the compressor, find the isentropic discharge $h_{2s}$ at $s_1$ and
the condenser pressure, then $h_2 = h_1 + (h_{2s} - h_1)/\eta_c$. State 3
is at the condenser pressure with the specified sub-cooling below
$T_{\text{cond}}$; read $h_3$. The throttle gives $h_4 = h_3$. Then
$\text{COP}_R = (h_1 - h_4)/(h_2 - h_1)$. With zero superheat, zero
sub-cooling, and $\eta_c = 1$ on the $-10\,\degC$ / $40\,\degC$ R-134a
case, the solver should return COP $\approx 4.1$, matching
\texttt{vapor\_compression.py}. Adding sub-cooling raises $q_c$ and the
COP; a compressor efficiency below one raises $h_2$ and lowers the COP.
The library \texttt{CoolProp} supplies the property look-ups for an
arbitrary refrigerant.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify the operating conditions (boiler pressure and temperature,
condenser pressure) for a $50\,\text{MW}$ Rankine plant that uses
cooling water at $20\,\degC$ as its heat sink. Target a thermal
efficiency of at least $35\,\%$.

\paragraph{Model-answer sketch.} Set the condenser first: with
$20\,\degC$ cooling water, a condenser saturation temperature near
$35\,\degC$ to $40\,\degC$ gives a workable approach; that fixes the
condenser pressure at roughly $6$ to $7\,\text{kPa}$ (the saturation
pressure of water at $\sim 38\,\degC$). To reach $35\,\%$ thermal
efficiency, the average temperature of heat addition must be high, so
choose a boiler at $\sim 10\,\text{MPa}$ and superheat to $\sim
500\,\degC$, a standard subcritical condition. The Carnot bound between
$500\,\degC$ ($773\,\si{\kelvin}$) and $38\,\degC$ ($311\,\si{\kelvin}$)
is $1 - 311/773 = 0.60$; a real Rankine plant reaching $\sim 0.65$ of
that lands near $39\,\%$, comfortably above the target. The
$50\,\text{MW}$ output then fixes the steam mass flow from $\dot W =
\dot m \, w_{\text{net}}$. State the assumed turbine efficiency
($\sim 0.85$) and note that reheat would be the next step to keep the
turbine-exit quality acceptable.
\end{exercise}

\begin{exercise}
Specify the pressure ratio and turbine inlet temperature for a
$10\,\text{MW}$ industrial Brayton-cycle gas turbine targeting a
specific work output of $300\,\text{kJ}/\text{kg}$ and a thermal
efficiency of at least $35\,\%$.

\paragraph{Model-answer sketch.} Two targets, two knobs ($r_p$ and
$T_3$). Efficiency depends mainly on $r_p$: $\eta_B = 1 -
r_p^{-0.286} \ge 0.35$ needs $r_p \ge 5$, but specific work for a given
$T_3$ peaks at an intermediate $r_p$, so pick a moderate $r_p \approx
12$ to $16$, which gives $\eta_B \approx 0.51$ to $0.54$ ideal (and
$\sim 0.35$ once compressor and turbine inefficiencies are applied).
With $T_1 = 300\,\si{\kelvin}$ and $r_p = 14$, $T_2 \approx 638\,\si{
\kelvin}$. The specific-work target $w_{\text{net}} = c_p[(T_3 - T_4) -
(T_2 - T_1)] = 300$ then sets $T_3$: solving with $T_4 = T_3
r_p^{-0.286}$ gives $T_3 \approx 1400$ to $1500\,\si{\kelvin}$, within
reach of cooled blades. The $10\,\text{MW}$ rating fixes the air mass
flow from $\dot m = \dot W/w_{\text{net}} = 10{,}000/300 \approx
33\,\text{kg}/\text{s}$. State the assumed component efficiencies
($\eta_c \approx 0.85$, $\eta_t \approx 0.88$).
\end{exercise}

\begin{exercise}
Specify the refrigerant, the evaporator temperature, and the condenser
temperature for a household refrigerator that maintains $4\,\degC$
inside in a $22\,\degC$ kitchen. State the COP target and the design
margin against the Carnot bound.

\paragraph{Model-answer sketch.} The evaporator must run below the
cold-space temperature and the condenser above the kitchen, so size the
heat exchangers for a pinch of about $10\,\si{\kelvin}$ on each side:
evaporator at $-6\,\degC$ ($267\,\si{\kelvin}$), condenser at
$32\,\degC$ ($305\,\si{\kelvin}$). The Carnot cooling COP for the
reservoir pair ($4\,\degC$ box, $22\,\degC$ kitchen) is $277/18 = 15.4$,
but the device works across the wider evaporator-to-condenser span, with
Carnot COP $267/38 = 7.0$. A realised COP near $0.5$ of that, so $\sim
3$ to $3.5$, is a sound target, leaving a healthy margin below the
bound. Choose a low-GWP natural refrigerant such as isobutane (R-600a)
or propane (R-290), which the \texttt{refrigerants.csv} table lists
among the low-GWP hydrocarbons used in domestic appliances, in a small
hermetic compressor. State the
$10\,\si{\kelvin}$ pinch assumptions explicitly, since they set the gap
between the reservoir Carnot COP and the device Carnot COP.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A combined-cycle plant's measured electrical output divided by the
fuel input gives a thermal efficiency of $40\,\%$, against a design
specification of $58\,\%$. Identify the most probable cause from
among (a) compressor degradation, (b) turbine blade damage, (c)
heat-recovery steam generator fouling, (d) condenser cooling-water
flow restriction, and state the diagnostic measurement that would
distinguish them.

\paragraph{Solution sketch.} An $18$-point efficiency shortfall is large
enough to point at a major component, and each candidate leaves a
distinct signature. (a) Compressor degradation shows as a fallen
pressure ratio and raised compressor-exit temperature for the same speed;
measure the compressor discharge pressure and temperature. (b) Turbine
blade damage shows as low turbine output for the measured inlet
conditions and possibly raised exhaust temperature; compare turbine
shaft power to the gas-path energy drop. (c) HRSG fouling shows as a
high gas-side exit temperature (heat not transferred to the steam) and
reduced steam production; measure the HRSG stack temperature and steam
flow. (d) Condenser flow restriction shows as a raised condenser
pressure and saturation temperature; measure the condenser vacuum. The
single most diagnostic reading is the gas-turbine exhaust and HRSG stack
temperatures: a normal gas turbine with a hot stack indicts the HRSG or
bottom cycle, while a low gas-turbine output with a normal stack indicts
the gas turbine itself.
\end{exercise}

\begin{exercise}
A residential air-conditioner uses noticeably more electricity in
the second half of the summer than the first half, while delivering
the same cooling. Identify the most probable cause from among (a)
refrigerant leak, (b) condenser fouling, (c) evaporator coil dirt,
(d) thermostat calibration drift, and the diagnostic measurement that
would identify each.

\paragraph{Solution sketch.} Same cooling for more electricity means a
fallen COP, so the suspect is anything that widens the
evaporator-to-condenser pressure span or starves the cycle. (a) A
refrigerant leak lowers the evaporator pressure and the suction
superheat rises; measure the low-side pressure and superheat. (b)
Condenser fouling (dirt on the outdoor coil, accumulated over the
summer, which fits the seasonal timing) raises the condenser pressure
and the compressor lift; measure the high-side pressure and the
condenser approach temperature. (c) Evaporator coil dirt or a clogged
filter restricts airflow, lowers the evaporator temperature, and can
ice the coil; measure the air-side temperature drop and pressure across
the indoor coil. (d) Thermostat drift would change the delivered
cooling, not the electricity per unit cooling, so it does not fit the
symptom. The seasonal onset most strongly indicts (b): an outdoor coil
collects dust and debris through a summer of operation, and a hose-down
is the first corrective step.
\end{exercise}

\begin{exercise}
A diesel engine's exhaust temperature is observed to be $50\,\degC$
higher than its commissioning value at the same load. Identify the
operational meaning and the corrective action.

\paragraph{Solution sketch.} A higher exhaust temperature at equal load
means combustion is finishing later in the cycle or more fuel is being
burned for the same output, both signs of degraded efficiency: heat that
should have been turned into work during expansion is leaving in the
exhaust. Probable causes are retarded injection timing, worn or fouled
injectors giving poor atomisation and late burn, reduced compression
(worn rings or leaking valves dropping the compression temperature), or
a fouled turbocharger or charge-air cooler raising the intake
temperature. Corrective actions, in order of intrusiveness: check and
clean or replace the air filter and charge-air cooler, verify and reset
injection timing, test and service the injectors, and (if those do not
recover the temperature) inspect compression and the valve train.
Rising exhaust temperature is a standard early-warning indicator on
marine and stationary diesels precisely because it integrates several
efficiency-eroding faults into one cheap measurement.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A 2023 patent application claims a heat pump that delivers a COP of
$10$ between a $20\,\degC$ indoor space and a $0\,\degC$ outdoor air
source. Identify the Carnot bound and write a single sentence
rejecting the claim.

\paragraph{Solution sketch.} The Carnot heating COP for these reservoirs
is $T_h/(T_h - T_c) = 293/(293 - 273) = 293/20 = 14.65$. A claimed COP
of $10$ is below this bound, so the claim does not violate the second
law outright; the careful rejection turns on the realised fraction.
``A COP of $10$ between these reservoirs would require a second-law
efficiency of $10/14.65 = 0.68$, far above the $0.4$ to $0.5$ that
production air-source heat pumps achieve across real heat exchangers and
a defrost duty, so the claim is implausible and the burden falls on the
applicant to show the operating boundary, the measurement protocol, and
any unaccounted input.'' This is the lesson of the two-step diagnosis:
when a claim clears the Carnot bound, the rejection must rest on the
realised-efficiency check, not on a bound it does not in fact exceed.
The companion \texttt{efficiency\_claim\_check.py} reports both the bound
and the implied fraction.
\end{exercise}

\begin{exercise}
A start-up advertises an internal-combustion engine that runs on
``water plus a catalyst,'' achieving $80\,\%$ thermal efficiency.
Identify the first-law violation and the diagnostic experiment that
would detect the hidden energy input.

\paragraph{Solution sketch.} Water is the product of hydrogen
combustion, not a fuel: it sits at the bottom of the energy ladder, with
no chemical energy left to release. A ``catalyst'' lowers a reaction
barrier but cannot supply energy, so a device that consumes only water
and a catalyst and produces net work is a first-kind perpetual motion
machine, forbidden by the first law. If the water is first electrolysed
to hydrogen and oxygen and the gases are then burned, the energy
recovered cannot exceed the electrical energy spent on electrolysis, so
the loop is energy-losing, not energy-producing. The diagnostic
experiment: draw a control boundary around the whole device and meter
every crossing, electrical input, any battery state of charge, any
consumable beyond water, over a run long enough to exhaust a hidden
store. The energy out will track an input the advertisement omitted (a
battery, a pressurised-gas cylinder, or a fuel additive); if the only
genuine input is water, the claimed output simply will not appear.
\end{exercise}

\begin{exercise}
A consumer-grade ``infrared sauna'' is sold as $99\,\%$ efficient. The
product converts $1\,\text{kW}$ of electricity into infrared radiation
that heats the user's body. Discuss what ``efficient'' means in this
context and whether the figure is consistent with thermodynamics.

\paragraph{Solution sketch.} Resistive heating converts electricity to
heat at very nearly $100\,\%$, so a $99\,\%$ figure for ``electricity to
heat'' is unremarkable and not the relevant quantity. ``Efficient'' for
a sauna must mean the fraction of the emitted heat that reaches the
intended target rather than warming the cabinet, the air, and the room.
That figure depends on geometry, emitter placement, and how much of the
radiation the body intercepts, and it is far below $99\,\%$ in practice
because much of the radiated and convected heat misses the user. The
$99\,\%$ claim is consistent with thermodynamics (it does not violate
any law) but it answers the wrong question: there is no Carnot bound on
a resistive heater, and quoting the conversion efficiency tells the
buyer nothing about delivered warmth. The honest figure is the
target-heating fraction, which the advertisement does not state.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer proposes to evaluate a power plant by ``how much
of the fuel ends up as useful electricity.'' Write a one-paragraph
reply that distinguishes first-law (energy) and second-law (exergy)
analyses and identifies when each is appropriate.

\paragraph{Model-answer sketch.} The proposed metric is the first-law
(energy) efficiency, electricity out over fuel energy in, and it is the
right number when the question is fuel burned per unit of delivered
work, for instance comparing two plants that both make only
electricity. It is silent, though, on the quality of the energy: a plant
that dumps half its fuel as $40\,\degC$ waste heat and one that dumps it
as $400\,\degC$ heat score the same on the first law, yet the second
plant has wasted far more work potential. The second-law (exergy)
analysis weights each energy stream by its capacity to do work, $1 -
T_0/T$ for heat at temperature $T$, and so reveals where the work
potential was destroyed (combustion, heat transfer across large
temperature gaps) rather than merely where energy left. Use the
first-law figure for fuel accounting and the exergy figure for finding
which component to redesign and for comparing plants that deliver heat
and power together.
\end{exercise}

\begin{exercise}
A facility manager argues that, since cogeneration uses ``waste
heat,'' it has zero marginal cost for the heating it provides. Write
a one-paragraph reply identifying the partial truth and the
qualifications.

\paragraph{Model-answer sketch.} The partial truth is real: when a power
cycle is already running to make electricity and would otherwise reject
its heat to a cooling tower or stack, capturing that heat does add
useful output for little extra fuel, so the marginal fuel cost of the
heat can indeed be small. The qualifications are several. First, raising
steam at a useful temperature for the heat load often forces the cycle
to run at a lower electrical efficiency (back-pressure operation), so
the heat is partly paid for in lost electricity, not in nothing.
Second, the recovered heat is delivered at a much lower temperature than
the combustion, so its exergy, its work value, is a fraction of its
energy; calling it free flatters it against electricity. Third, the
plant carries capital, maintenance, and distribution costs for the heat
network that are not zero. The right statement is that cogeneration has
a low marginal fuel cost for heat when the heat load matches the cycle's
rejection in temperature and timing, not that the heat is free.
\end{exercise}

\begin{exercise}
A potential customer asks why a heat pump for home heating is
preferable to a high-efficiency natural-gas furnace. Write a one-
paragraph reply that uses the Carnot bound and the typical realised
efficiencies to frame the comparison.

\paragraph{Model-answer sketch.} A condensing gas furnace converts close
to all of the fuel's heating value into delivered heat, but it cannot
exceed $100\,\%$: each joule of gas yields at most a joule of heat. A
heat pump is not bound by $100\,\%$ because it moves heat rather than
making it; its Carnot heating COP between, say, a $20\,\degC$ indoor and
a $0\,\degC$ outdoor is $293/20 = 14.7$, and a real unit delivers $3$ to
$4$ units of heat per unit of electricity. So one unit of electricity
through a heat pump delivers three to four units of heat, while one unit
of gas through a furnace delivers at most one. The comparison that
matters to the customer then turns on the local prices of electricity
and gas and on the carbon intensity of the grid: the heat pump wins on
delivered heat per unit of purchased energy by a wide margin, and the
gap narrows only where electricity is far costlier than gas or where the
outdoor temperature is so low that the COP collapses toward one.
\end{exercise}
