\chapter{Engineering quantum mechanics}
\label{vol04:ch12}

\epigraph{It is safe to say that no one understands quantum mechanics.
What I am going to tell you about is what nature behaves like.}{Richard
Feynman, \emph{The Character of Physical Law} (1965), Lecture~6 on
probability and uncertainty, paraphrased in the form widely cited; the
working point for this chapter is the second sentence, which states the
engineering operational stance.}

\chapmeta{Half-life: durable. Archetypes: scaling, interface. Exercise target: 26. Project track: bench measurement plus analysis (hybrid). Hazard class: low (low-power diodes, LEDs, photodiodes, Class~1 or Class~2 lasers only). Reader path default: standard, with sections explicitly tagged. Named cases: none directly; the chapter's failure section treats the quantum-noise floor as a fundamental limit rather than an accident.}

\noindent
Quantum mechanics enters engineering at the scale where the action of
the relevant process approaches Planck's constant. The chapter treats
the working quantum content the practising engineer carries:
energy-band semiconductors, photon-electron interactions in
LED and laser diodes and photovoltaic cells, the MOSFET transistor as
a quantum-mechanical device, MRI, electron microscopy, X-ray, and the
shot-noise and Johnson-noise floors that set the limits of classical
electronics. The full quantum-mechanical apparatus is treated in
Volume V (where it serves as the substrate for chemistry); here we
extract the engineering content.

\section{Why classical physics is not enough at the device scale}\pathtag{core}

A device with characteristic length $L$, characteristic energy $E$,
and characteristic time $\tau$ has \engterm{characteristic action}
$S \sim E\tau$. When $S$ is much larger than $\hbar = 1.055\times
10^{-34}\,\si{\joule\second}$, classical physics is the correct
description. When $S$ approaches $\hbar$, quantum effects become
operationally important; when $S$ is on the order of $\hbar$, the
device's behaviour is dominated by quantum mechanics.

For an electron in a copper wire at room temperature, the thermal
energy is $k_{B}T \approx 4\times 10^{-21}\,\si{\joule}$; the
relaxation time is $\tau \approx 25\,\si{\femto\second}$; the
action is $S \sim 10^{-34}\,\si{\joule\second} \sim \hbar$. The
electron is quantum-mechanical at the atomic scale. A wire of
macroscopic length contains $\sim 10^{22}$ electrons; the
ensemble-averaged behaviour (Ohm's law $V = IR$) is classical even
though each carrier is quantum. The classical-quantum boundary is at
the device's smallest feature size, not at its macroscopic dimension.

Three operational regimes follow:
\begin{itemize}[leftmargin=*]
  \item \engterm{Classical regime}: feature size $\gg$ de~Broglie
    wavelength. Mechanical engineering, structural engineering, most
    fluid mechanics, most circuit design above the nanometre level.
  \item \engterm{Mesoscopic regime}: feature size comparable to the
    de~Broglie wavelength or the coherence length. Modern transistors
    (channel length below $10\,\si{\nano\meter}$), quantum dots,
    single-photon devices, low-temperature superconducting circuits.
  \item \engterm{Quantum regime}: feature size $\ll$ de~Broglie
    wavelength, action $\sim \hbar$. Individual atomic and molecular
    processes; MRI; electron microscopy; X-ray sources.
\end{itemize}

The \engterm{de~Broglie wavelength} of a particle of momentum $p$ is
\[
\lambda_{\text{dB}} = \frac{h}{p},
\]
where $h = 2\pi\hbar$. For a thermal electron at room temperature
($E = (3/2)k_{B}T \approx 0.04\,\si{\electronvolt}$, $p =
\sqrt{2m_{e}E} \approx 1.1\times 10^{-25}\,\si{\kilogram\meter\per
\second}$), $\lambda_{\text{dB}} \approx 6\,\si{\nano\meter}$. Modern
transistors with channel lengths below this scale are unambiguously
in the mesoscopic regime; their design depends on quantum-mechanical
calculation rather than on classical drift-diffusion.

\begin{mastery}
The action criterion has a sharper statement as the Heisenberg
uncertainty principle, and the principle is an engineering bound, not a
philosophical remark. Position and momentum cannot both be sharp:
$\Delta x\,\Delta p \gtrsim \hbar/2$. The bound sets the smallest box a
carrier can be confined in before its zero-point motion matters. Squeeze
an electron into a region $\Delta x$ and it acquires a momentum spread
$\Delta p \gtrsim \hbar/2\Delta x$ and so a confinement kinetic energy of
order $(\Delta p)^{2}/2m^{*} \sim \hbar^{2}/(8m^{*}\Delta x^{2})$, the
same scale as the particle-in-a-box ground state and the reason a thinner
quantum well emits at higher energy. For an electron of conduction-band
mass $m^{*} = 0.26\,m_e$ confined to $\Delta x = 5\,\si{\nano\meter}$, the
confinement energy is of order $\hbar^{2}/(8m^{*}\Delta x^{2}) \approx
0.0015\,\si{\electronvolt}$ as an uncertainty-principle floor, with the
particle-in-a-box ground state more than an order of magnitude above it and
of order $k_B T$ at room temperature, which is why a sub-ten-nanometre
transistor channel cannot be treated as a classical sheet of charge. The energy-time form $\Delta E\,\Delta t \gtrsim \hbar/2$
is the bound an engineer meets in the linewidth of a laser and the
ring-down of a resonator: a state that lives for a time $\tau$ has an
energy width at least $\hbar/2\tau$, so a transition with a $1\,\si{\nano
\second}$ lifetime has a natural linewidth no narrower than about
$0.1\,\si{\micro\electronvolt}$, the floor below which no amount of optical
engineering narrows the line. The uncertainty principle is the quantitative
content of the action criterion that opened the chapter: it states, in
energy and length, exactly where the classical description must give way.
\end{mastery}

Two experimental results forced quantisation on physics before any
device used it, and both remain the working starting points for the
engineer. The photoelectric effect shows that light delivers energy in
quanta $h\nu$: below a threshold frequency no electrons are emitted at
any intensity, and above it the maximum electron kinetic energy rises
linearly with frequency at a slope fixed by Planck's constant
(Figure~\ref{fig:vol04ch12:photoelectric-stopping}). The blackbody
spectrum shows that the electromagnetic field's modes are quantised:
the measured radiance peaks and then falls toward short wavelength
(Figure~\ref{fig:vol04ch12:blackbody-curves}), where the classical
equipartition result diverges. The code in
\texttt{code/photon\_energy.py} carries the photon arithmetic and the
$1.24\,\text{eV}\cdot\si{\micro\meter}$ rule; the data in
\texttt{data/photoelectric\_stopping.csv} and
\texttt{data/blackbody\_spectrum.csv} reproduce the two figures.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-photoelectric-stopping.tex}

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-blackbody-curves.tex}

\subsection{The Planck law from mode-counting and Bose statistics}\pathtag{mastery}

The blackbody curve is the first place quantisation became unavoidable,
and the derivation is short enough to carry in full because it isolates
the single quantum assumption that fixes the entire photonics of the
chapter. A cavity of volume $V = L^{3}$ holds the electromagnetic field.
The field's standing-wave modes are the same count performed for the
electron in a box: allowed wavevectors form a lattice of points spaced
$\pi/L$, and the number of modes with wavevector magnitude below $k$,
counting the two polarisations of light and the positive octant, is
\[
N(k) = 2\cdot\frac{1}{8}\cdot\frac{(4/3)\pi k^{3}}{(\pi/L)^{3}}
     = \frac{V k^{3}}{3\pi^{2}}.
\]
Light has $\omega = ck$, so changing variable to frequency
$\nu = \omega/2\pi = ck/2\pi$ and differentiating gives the number of
modes per unit volume per unit frequency,
\[
g(\nu) = \frac{1}{V}\frac{dN}{d\nu} = \frac{8\pi\nu^{2}}{c^{3}},
\]
the mode density that the classical and the quantum calculation share.
The two calculations part at the energy assigned to each mode. The
classical equipartition theorem gives every mode the same average energy
$k_B T$, and the product
\[
u_{\text{RJ}}(\nu) = g(\nu)\,k_B T = \frac{8\pi\nu^{2}}{c^{3}}\,k_B T
\]
is the Rayleigh-Jeans law, which rises without bound as $\nu^{2}$. The
integral over all frequencies diverges, the ultraviolet catastrophe of
Figure~\ref{fig:vol04ch12:planck-vs-rj}: a cavity in equilibrium would
hold infinite energy. Planck's resolution is to treat each mode as a
harmonic oscillator that can hold energy only in quanta $h\nu$, so the
mode energy is the integer $n$ times $h\nu$, weighted by the Boltzmann
probability $\exp(-nh\nu/k_B T)$. Summing the geometric series for the
mean,
\[
\langle E\rangle
= \frac{\sum_{n=0}^{\infty} n h\nu\,e^{-nh\nu/k_B T}}
       {\sum_{n=0}^{\infty} e^{-nh\nu/k_B T}}
= \frac{h\nu}{e^{h\nu/k_B T} - 1},
\]
the Bose-Einstein mean occupation of the mode times its quantum of
energy. The spectral energy density is the mode density times this mean
energy,
\[
u(\nu)\,d\nu = \frac{8\pi\nu^{2}}{c^{3}}\,
\frac{h\nu}{e^{h\nu/k_B T} - 1}\,d\nu,
\]
the Planck law. Two checks tie it to the rest of the chapter. At low
frequency $h\nu \ll k_B T$ the denominator expands to $h\nu/k_B T$, the
quantum of energy cancels, and the Bose mean returns the equipartition
$k_B T$, so Planck reduces to Rayleigh-Jeans where the quantum is small,
the agreement at small $x$ in Figure~\ref{fig:vol04ch12:planck-vs-rj}.
At high frequency the exponential dominates and the spectrum falls as
$e^{-h\nu/k_B T}$, killing the catastrophe. The peak, found by setting
the derivative to zero in the dimensionless variable $x = h\nu/k_B T$,
sits at $x \approx 2.82$, which is Wien's displacement law $\lambda_{\max}
T = b$ used in the filament worked example; and integrating $u(\nu)$ over
all frequencies returns the Stefan-Boltzmann $\sigma T^{4}$ used wherever
this book balances radiated power. The single change from $k_B T$ per mode
to the Bose mean energy per mode is the whole of the quantum hypothesis at
the level an engineer applies it, and the same Bose occupation reappears as
the photon chemical potential $\mu = eV$ that set the detailed-balance
ceiling of the solar-cell case study.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-planck-vs-rj.tex}

\section{Photon, electron, semiconductor}\pathtag{core}

A \engterm{photon} is the quantum of the electromagnetic field. Its
energy is $E = h\nu = \hbar\omega$, where $\nu$ is the frequency. At
$\lambda = 633\,\si{\nano\meter}$ (red HeNe laser), $E = 1.96\,\si{
\electronvolt}$. At $\lambda = 1\,\si{\micro\meter}$ (near-infrared),
$E = 1.24\,\si{\electronvolt}$. The simple rule $E[\si{\electronvolt}]
= 1.24/\lambda[\si{\micro\meter}]$ is the working conversion the
engineer carries to bench experiments.

An \engterm{electron} in vacuum has rest mass $m_{e} = 9.109\times
10^{-31}\,\si{\kilogram}$ and charge $-e$. In a solid, the electron
moves through a periodic potential created by the atomic lattice; its
behaviour is described by Bloch's theorem, which gives the
wavefunction as a plane-wave envelope times a lattice-periodic
function. The effective mass $m^{*}$ describes the electron's
response to slowly-varying external forces and differs from the
free-electron mass: in silicon's conduction band, $m^{*}_{e} \approx
0.26\,m_{e}$; in GaAs, $m^{*}_{e} \approx 0.067\,m_{e}$. Effective
mass is the working content of every band-structure calculation in
engineering.

\begin{mastery}
The square-root density of states $g(E) \propto \sqrt{E - E_{c}}$ of
Figure~\ref{fig:vol04ch12:density-of-states} is the three-dimensional
result. Confining the carriers to a thin layer changes it. In a
quantum well (one dimension confined) the density of states becomes a
staircase of constant steps; in a quantum wire (two dimensions
confined) it acquires inverse-square-root spikes; in a quantum dot
(all three confined) it collapses to a set of delta functions, an
artificial atom. The design payoff is sharper gain spectra: the
quantum-well laser concentrates the available states near the band
edge, lowering threshold current and narrowing the emission line
relative to a bulk device. Quantum-dot films exploit the discrete
levels for narrow, size-tunable emission in displays. The same
confinement that raises the ground-state energy of a particle in a box
(treated in the exercises) sets the emission wavelength here.
\end{mastery}

A \engterm{semiconductor} is a material with a small energy gap
between its valence and conduction bands. At $T = 0$ the valence band
is fully occupied and the conduction band is empty; the material is
an insulator. At finite temperature, thermal excitation lifts some
electrons across the gap, leaving holes (effective positive charges)
in the valence band. The intrinsic carrier density is
\[
n_{i} = \sqrt{N_{c}N_{v}}\exp(-E_{g}/2k_{B}T),
\]
where $N_{c}$, $N_{v}$ are effective density of states in the
conduction and valence bands, and $E_{g}$ is the band gap. For
silicon at $300\,\si{\kelvin}$, $E_{g} = 1.12\,\si{\electronvolt}$
and $n_{i} \approx 10^{10}\,\si{\per\centi\meter\cubed}$; for GaAs,
$n_{i} \approx 2\times 10^{6}\,\si{\per\centi\meter\cubed}$; for
germanium, $n_{i} \approx 2\times 10^{13}\,\si{\per\centi\meter
\cubed}$. The band gap is the property that places a material in one
of three electrical classes
(Figure~\ref{fig:vol04ch12:energy-bands}); the band gaps, gap types,
effective masses, and intrinsic carrier densities used through this
chapter are collected in \texttt{data/semiconductor\_properties.csv}.
The reference work for these numbers and for the device physics that
follows is Sze and Ng \cite{text:sze-semiconductor}.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-energy-bands.tex}

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-density-of-states.tex}

\subsection{From the particle in a box to the band-edge carrier
density}\pathtag{mastery}

The intrinsic-carrier formula carries two objects that the chapter so
far has stated rather than derived: the effective density of states
$N_c$, $N_v$, and the Boltzmann occupation that places carriers in the
band. Both follow from one calculation, the particle in a box, and the
result is the working bridge between the quantum picture and the device
arithmetic.

A conduction electron near the band edge behaves as a free particle of
mass $m^{*}$ confined to the crystal, a box of side $L$ with volume
$V = L^{3}$. The stationary states of a particle in a one-dimensional
box of width $L$ with infinite walls are the standing waves that vanish
at both ends, $\psi_n(x) \propto \sin(n\pi x/L)$, with quantised
wavenumber $k_n = n\pi/L$ and energy
\[
E_n = \frac{\hbar^{2}k_n^{2}}{2m^{*}}
     = \frac{n^{2}\pi^{2}\hbar^{2}}{2m^{*}L^{2}},
\qquad n = 1, 2, 3, \dots
\]
This is the same ladder that sets the quantum-well emission wavelength
in the worked examples; here we let the box be macroscopic and count the
states. In three dimensions the allowed wavevectors form a lattice of
points spaced $\pi/L$ along each axis, so each state occupies a volume
$(\pi/L)^{3}$ of the positive octant of $k$-space. The number of states
with wavevector magnitude below $k$, counting the two spin orientations
and the single octant, is
\[
N(k) = 2 \cdot \frac{1}{8}\cdot\frac{(4/3)\pi k^{3}}{(\pi/L)^{3}}
     = \frac{V k^{3}}{3\pi^{2}}.
\]
Differentiating and changing variable from $k$ to energy through
$E = \hbar^{2}k^{2}/2m^{*}$ gives the \engterm{density of states} per
unit volume per unit energy at the band edge,
\[
g(E) = \frac{1}{V}\frac{dN}{dE}
     = \frac{1}{2\pi^{2}}\left(\frac{2m^{*}}{\hbar^{2}}\right)^{3/2}
       \sqrt{E - E_c},
\]
the square-root edge of Figure~\ref{fig:vol04ch12:density-of-states}.
The carriers in this band are electrons, so the probability that a state
of energy $E$ is filled is the Fermi-Dirac function,
\[
f(E) = \frac{1}{1 + \exp\!\big((E - E_F)/k_B T\big)},
\]
plotted in Figure~\ref{fig:vol04ch12:fermi-dirac}. At zero temperature
$f$ is a step: every state below the Fermi level $E_F$ is full and every
state above is empty. Finite temperature softens the step over a width
of a few $k_B T$, and the conduction band picks up the thermally excited
tail.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-fermi-dirac.tex}

The electron density in the conduction band is the density of states
weighted by the occupation, integrated upward from the band edge,
\[
n = \int_{E_c}^{\infty} g(E)\,f(E)\,dE.
\]
In a non-degenerate semiconductor the Fermi level sits well below the
conduction edge, $E_c - E_F \gg k_B T$, so for every state in the band
the exponential in $f$ dominates the $+1$ and the Fermi-Dirac function
collapses to the Boltzmann tail $f(E) \approx \exp(-(E - E_F)/k_B T)$.
The integral then separates into a constant times an exponential,
\[
n = N_c\,\exp\!\big(-(E_c - E_F)/k_B T\big),
\qquad
N_c = 2\left(\frac{m^{*}_e k_B T}{2\pi\hbar^{2}}\right)^{3/2},
\]
and the integral over the square-root density of states against the
Boltzmann factor is what produces the $T^{3/2}$ prefactor $N_c$, the
\engterm{effective density of states} that the intrinsic-carrier formula
carried as a symbol. The valence band gives the hole density $p = N_v
\exp(-(E_F - E_v)/k_B T)$ by the same argument. Multiplying the two,
the Fermi level cancels and the product is fixed by the gap alone,
\[
np = N_c N_v\,\exp(-E_g/k_B T) \equiv n_i^{2},
\]
the mass-action law, and setting $n = p = n_i$ in the intrinsic case
recovers $n_i = \sqrt{N_c N_v}\exp(-E_g/2k_B T)$, the formula stated
above. The Boltzmann step is accurate to better than a percent for
$E_c - E_F \gtrsim 3k_B T$ and fails inside a degenerate (heavily doped)
semiconductor, where the Fermi level enters the band and the full
Fermi-Dirac integral is required. The engineering payoff is that one
particle-in-a-box count, weighted by Fermi-Dirac statistics in the
non-degenerate limit, delivers both the $\sqrt{E}$ density of states and
the exponential carrier law that the rest of the chapter uses without
re-derivation.

\engterm{Doping} replaces a fraction of the host atoms with impurities
of higher (n-type, e.g., phosphorus in silicon) or lower (p-type,
boron in silicon) valence. A doping density of $10^{18}\,\si{\per
\centi\meter\cubed}$ in silicon produces a free-electron density of
the same value, eight orders of magnitude above the intrinsic
density. The free-electron concentration is now set by the dopant
rather than by the thermal excitation; the carrier density is
\engterm{extrinsic}.

A \engterm{pn~junction} forms at the boundary between an n-doped
region and a p-doped region. The chemical-potential difference drives
electron diffusion from n to p and hole diffusion from p to n; the
charged regions left behind (positive ionised donors on the n side,
negative ionised acceptors on the p side) build up an electric field
opposing further diffusion. At equilibrium the diffusion and drift
currents cancel and a depletion region of width $\sim 100\,\si{\nano
\meter}$ at typical doping forms. The band edges bend across the
junction by the built-in potential while the Fermi level stays flat
(Figure~\ref{fig:vol04ch12:pn-junction-bands}); this single bent-band
picture, driven electrically or illuminated, is the object behind the
diode, the LED, the laser, and the solar cell.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-pn-junction-bands.tex}

The built-in potential is the equilibrium band bending,
\[
V_{bi} = \frac{k_{B}T}{e}\,\ln\!\left(\frac{N_{A}N_{D}}{n_{i}^{2}}
\right),
\]
which for silicon with $N_{A} = N_{D} = 10^{17}\,\si{\per\centi\meter
\cubed}$ and $n_{i} = 10^{10}\,\si{\per\centi\meter\cubed}$ evaluates to
$V_{bi} = 0.0259\,\ln(10^{34}/10^{20}) = 0.0259 \times 32.2 = 0.83\,\si{
\volt}$. The depletion width at zero bias follows from solving Poisson's
equation across the space-charge region,
\[
W = \sqrt{\frac{2\varepsilon V_{bi}}{e}\left(\frac{1}{N_{A}} +
\frac{1}{N_{D}}\right)},
\]
which for these dopings and silicon's permittivity gives $W \approx
0.15\,\si{\micro\meter}$, the order claimed above. The width narrows
under forward bias (the applied voltage subtracts from $V_{bi}$) and
widens under reverse bias, the modulation that makes the junction a
voltage-controlled capacitor and the basis of the varactor.

Both formulas come from one electrostatics calculation worth carrying
once, because every junction device in the chapter inherits it. The
\engterm{depletion approximation} idealises the space-charge region as
fully stripped of mobile carriers and the neutral bulk as perfectly
screened, so the charge density is a step: a slab of ionised acceptors
$\rho = -eN_A$ on the p side from $-x_p$ to $0$, and a slab of ionised
donors $\rho = +eN_D$ on the n side from $0$ to $x_n$. Charge neutrality
across the whole junction requires the two slabs to carry equal and
opposite charge, $N_A x_p = N_D x_n$, so the depletion region pushes
farther into the more lightly doped side. Poisson's equation in one
dimension, $d^{2}\phi/dx^{2} = -\rho/\varepsilon$, integrates once to the
field. On the n side,
\[
E(x) = -\frac{d\phi}{dx}
     = \frac{eN_D}{\varepsilon}(x - x_n),
\]
a triangle that is zero at the neutral edge $x_n$ and grows linearly
toward the junction; the p side mirrors it, and the two meet at a peak
field $E_{\max} = eN_D x_n/\varepsilon = eN_A x_p/\varepsilon$ at $x = 0$.
Integrating the field once more gives the potential as two parabolas, and
the total band bending is the area under the triangular field profile, the
area of a triangle of base $W = x_p + x_n$ and height $E_{\max}$,
\[
V_{bi} = \frac{1}{2}E_{\max}\,W
       = \frac{e}{2\varepsilon}\,
         \frac{N_A N_D}{N_A + N_D}\,W^{2}.
\]
Solving for the width returns the depletion-width formula above with
$1/N_A + 1/N_D$ in place of the product-over-sum, the same expression
rearranged. The built-in potential itself is the equilibrium band
bending that flattens the Fermi level across the junction, and it is fixed
by the same mass-action law derived earlier: the n-side and p-side
electron densities differ by the Boltzmann factor $\exp(eV_{bi}/k_B T)$,
which inverts to the logarithm $V_{bi} = (k_B T/e)\ln(N_A N_D/n_i^{2})$
stated above. Two device facts read straight off this calculation. The
peak field scales as the square root of doping, so a heavily doped
junction reaches the avalanche-breakdown field at a lower reverse voltage,
the trade behind every diode's voltage rating. The depletion charge per
unit area is $Q = eN_D x_n$ and the junction capacitance is $C = \varepsilon
/W$, which falls as the reverse bias widens $W$; the capacitance-voltage
slope is the standard route to the doping profile, and the bias-tunable
capacitance is the varactor exploited in voltage-controlled oscillators.

Applying a voltage $V$ across the junction with the p side at higher
potential than the n side (\engterm{forward bias}) reduces the
barrier and exponentially increases the current; reversing the bias
slightly enlarges the barrier and the current saturates at a small
reverse-saturation value. The Shockley diode equation is
\[
I(V) = I_{s}\left[\exp(eV/k_{B}T) - 1\right],
\]
with $I_{s}$ the reverse-saturation current (typically $10^{-12}\,\si{
\ampere}$ for a small-signal silicon diode at room temperature). The
$26\,\si{\milli\volt}$ thermal voltage $k_{B}T/e$ at $300\,\si{\kelvin}$
is the working number for diode-circuit analysis; a $60\,\si{\milli
\volt}$-per-decade increase in current is the standard diode-curve
slope.

\section{LED, laser, photovoltaic cell}\pathtag{core}

The pn junction emits and absorbs photons at the band-gap energy
when electrons and holes recombine or are generated.

A \engterm{light-emitting diode} (LED) is a pn junction in a direct-
band-gap semiconductor, forward-biased. Electrons and holes injected
across the junction recombine radiatively, emitting photons at energy
$\hbar\omega \approx E_{g}$. Red LEDs use AlGaAs or AlGaInP
($E_{g} \approx 1.9\,\si{\electronvolt}$); green LEDs use GaP or
InGaN at modest indium content; blue and white LEDs use InGaN at
high indium content. Silicon and germanium are indirect-band-gap
semiconductors and do not emit light efficiently as LEDs; they are
not used for emission. The internal quantum efficiency of modern
LEDs exceeds $80\,\%$; the external efficiency (limited by light
extraction from the high-index semiconductor) is in the $40$ to
$70\,\%$ range with good encapsulation.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-led-laser-levels.tex}

A \engterm{laser diode} differs from an LED in that an optical cavity
is formed by cleaved end facets, which provide the feedback for
stimulated emission
(Figure~\ref{fig:vol04ch12:led-laser-levels}). The standard reference
for the photonics in this section is Saleh and Teich
\cite{text:saleh-teich-photonics}. Above the threshold current, stimulated emission
dominates spontaneous emission, producing coherent monochromatic
output. Modern telecom laser diodes at $1.55\,\si{\micro\meter}$ have
threshold currents of $\sim 20\,\si{\milli\ampere}$, output powers of
$10\,\si{\milli\watt}$ continuous, and linewidths below $10\,\si{
\mega\hertz}$ with distributed-feedback structures.

A \engterm{photovoltaic cell} (solar cell) is a pn junction in a
semiconductor, reverse-biased or in open circuit, illuminated. Photons
of energy above $E_{g}$ create electron-hole pairs in the depletion
region; the built-in field separates them, sending electrons to the
n side and holes to the p side, producing a photocurrent. The
short-circuit current scales with photon flux; the open-circuit
voltage approaches $E_{g}/e$ at low temperature and high illumination.

The Shockley-Queisser limit for a single-junction solar cell at
standard $1\,\text{sun}$ illumination (AM1.5G, $1000\,\si{\watt\per
\meter\squared}$) is $\sim 33\,\%$ at the optimum band gap of
$1.34\,\si{\electronvolt}$ (close to GaAs). Modern silicon cells at
$E_{g} = 1.12\,\si{\electronvolt}$ achieve $\sim 26\,\%$ in
laboratory record cells (current as of 2024); commercial silicon
modules reach $20$ to $23\,\%$. Multi-junction cells stacking
materials of different band gaps reach $\sim 47\,\%$ at concentrated
illumination; the cost is the additional materials and the
mechanical complexity of the concentrator system. The single-junction
ceiling and the location of its optimum band gap are shown in
Figure~\ref{fig:vol04ch12:shockley-queisser}.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-shockley-queisser.tex}

\begin{archetype}[Scaling and interface]
The LED, the laser diode, and the photovoltaic cell are the same pn
junction operated in three different regimes. Each is the interface
archetype made explicit: at the junction, electrons and holes
recombine to produce light or are separated by the built-in field to
produce current. The chapter's working content is that the device's
behaviour is set by the physics of the junction, scaled by the area
of the device and the temperature of operation.
\end{archetype}

\subsection{Diffusion, the Einstein relation, and the length that sets the
saturation current}\pathtag{standard}

The Shockley saturation current $I_s$ has appeared as a symbol in the diode
law, the LED, and the solar cell, and the case studies charged their voltage
losses to it, but the number itself comes from one length that ties a
material's carrier lifetime to its diode current. When forward bias raises
the minority-carrier density at the depletion edge by the Boltzmann factor
$\exp(eV/k_BT)$, the excess carriers do not sit there; they diffuse into the
neutral bulk and recombine, and the current the junction carries is the rate
at which they are supplied to that diffusion.

Two transport processes act on the carriers, and both are governed by one
coefficient. Diffusion drives carriers down a concentration gradient with a
flux $-D\,dn/dx$, the diffusion constant $D$ measuring how fast they spread;
drift drives them along an electric field with a velocity $\mu E$, the
mobility $\mu$ measuring how fast the field pushes them. The two are not
independent, because in thermal equilibrium the diffusion that spreads a
carrier cloud must exactly balance the drift that a built-in field imposes
on it, and requiring that balance gives the \engterm{Einstein relation}
\[
\frac{D}{\mu} = \frac{k_BT}{e} = V_T,
\]
the same thermal voltage that sets the diode slope. A material's mobility,
which a Hall measurement returns, therefore fixes its diffusion constant at
once: for silicon electrons with $\mu_n = 0.15\,\si{\meter\squared\per\volt
\per\second}$, the Einstein relation gives $D_n = 0.0259\times 0.15 =
3.9\times 10^{-3}\,\si{\meter\squared\per\second} = 39\,\si{\centi\meter
\squared\per\second}$.

The excess carriers injected at the edge decay into the bulk as they
recombine, and the steady diffusion equation $D\,d^{2}\Delta n/dx^{2} =
\Delta n/\tau$ has the exponential solution $\Delta n(x) = \Delta n(0)\,
\exp(-x/L)$ of Figure~\ref{fig:vol04ch12:diffusion-profile}, with the
\engterm{diffusion length}
\[
L = \sqrt{D\tau},
\]
the distance a minority carrier travels before it recombines. For the
silicon electrons above with a lifetime $\tau = 10\,\si{\micro\second}$,
$L_n = \sqrt{(39)(10\times 10^{-6})} = 0.020\,\si{\centi\meter} =
200\,\si{\micro\meter}$, comparable to a wafer thickness, which is why a
silicon solar cell can collect carriers generated deep in its bulk. The
saturation current is the diffusion current evaluated at the edge, the slope
of the profile,
\[
I_s = eA\left(\frac{D_n n_{p0}}{L_n} + \frac{D_p p_{n0}}{L_p}\right),
\]
so a longer diffusion length lowers $I_s$ and, through the open-circuit
voltage $V_{oc} = V_T\ln(J_{sc}/J_0)$, raises the cell's voltage. The chain
is now closed: the mobility a Hall bar measures fixes the diffusion constant
through the Einstein relation, the diffusion constant and the lifetime fix
the diffusion length, and the diffusion length fixes the saturation current
that every junction device in the chapter carried as a symbol.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-diffusion-profile.tex}

\section{An end-to-end case study: a silicon photovoltaic cell}\pathtag{standard}

The chapter's apparatus pays off when an engineer must judge a real
device from first principles, the spectrum it sees, and a handful of
material numbers. We take a single-junction monocrystalline silicon
solar cell and carry it from the solar spectrum through the band-gap
cutoff, the detailed-balance ceiling, the open-circuit voltage,
short-circuit current, and fill factor, then the real losses that
separate a record cell from a fielded module, ending in a verdict an
engineer would sign. The cell is a good vehicle because every move of
the LED-laser-photovoltaic section appears in it: the band gap as
absorber threshold, the photon flux as current, the junction voltage
as the diode law, and the area scaling that turns a current density
into a panel rating.

\subsection{The spectrum and the band-gap cutoff}

The standard terrestrial test condition is one sun of AM1.5G
illumination, a total irradiance of $1000\,\si{\watt\per\meter
\squared}$ integrated over the spectrum of
Figure~\ref{fig:vol04ch12:am15-spectrum}. Silicon has a band gap
$E_{g} = 1.12\,\si{\electronvolt}$. The working rule
$\lambda[\si{\micro\meter}] = 1.24/E[\si{\electronvolt}]$ places the
absorption edge at $\lambda_{g} = 1.24/1.12 = 1.107\,\si{\micro
\meter}$. Photons of longer wavelength carry less than the gap energy
and pass through the cell unabsorbed; photons of shorter wavelength are
absorbed, but each one creates a single electron-hole pair and the
energy it carries above the gap is lost as heat as the carrier relaxes
to the band edge. These are the two single-junction losses, and they
pull the optimum gap in opposite directions.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-am15-spectrum.tex}

The fraction of the solar power that lies above the silicon gap is
about $77\,\%$; the rest is in the infrared tail to the right of the
cutoff in Figure~\ref{fig:vol04ch12:am15-spectrum} and cannot be used
by silicon at all. Of the absorbed power, the thermalisation of the
excess photon energy removes roughly another third, leaving on the
order of half the incident power available as electron-hole pairs at
the band edge before any junction physics enters. The running fraction
is tracked in the waterfall of
Figure~\ref{fig:vol04ch12:sc-cell-loss-waterfall}.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-sc-cell-loss-waterfall.tex}

\subsection{Short-circuit current}

Each above-gap photon that is absorbed contributes one unit of charge
to the photocurrent. The AM1.5G photon flux above the silicon gap is
about $\Phi \approx 2.7\times 10^{21}\,\si{\per\meter\squared\per
\second}$ (the integral of the spectrum, converted from energy to
photon count, out to $1107\,\si{\nano\meter}$). The maximum
short-circuit current density is then
\[
J_{sc} = e\,\Phi \approx (1.6\times 10^{-19})(2.7\times 10^{21})
\approx 43\,\si{\milli\ampere\per\centi\meter\squared},
\]
the value a defect-free silicon cell that collected every above-gap
photon would deliver. Record silicon cells reach $42\,\si{\milli
\ampere\per\centi\meter\squared}$, within a few percent of this
ceiling, because silicon's indirect gap forces a thick (about
$200\,\si{\micro\meter}$) absorber and good light trapping to catch the
weakly-absorbed red and near-infrared photons. The short-circuit
current scales directly with illumination: at the concentration ratio
$C$ used in concentrator systems, $J_{sc}$ rises by the factor $C$.

\subsection{Open-circuit voltage}

The junction is a diode, and in the light it carries both the
photocurrent and the dark diode current. The illuminated current is
\[
J(V) = J_{sc} - J_{0}\left[\exp(eV/k_{B}T) - 1\right],
\]
the superposition of the Shockley diode law and the constant
photocurrent. At open circuit $J = 0$, so
\[
V_{oc} = \frac{k_{B}T}{e}\,\ln\!\left(\frac{J_{sc}}{J_{0}} + 1\right).
\]
The saturation current density $J_{0}$ carries the recombination
physics; the detailed-balance ideal sets a radiative floor of order
$J_{0} \approx 10^{-15}\,\si{\milli\ampere\per\centi\meter\squared}$
for silicon. With $J_{sc} = 43\,\si{\milli\ampere\per\centi\meter
\squared}$ and $k_{B}T/e = 0.0259\,\si{\volt}$, the radiative-limit
voltage is
\[
V_{oc} \approx 0.0259\,\ln\!\left(\frac{43}{10^{-15}}\right)
= 0.0259 \times \ln(4.3\times 10^{16}) = 0.0259 \times 38.3
= 0.99\,\si{\volt}.
\]
The gap is $1.12\,\si{\volt}$ in energy units, so even the perfect cell
loses about $0.13\,\si{\volt}$ to the entropy of the second law: a
finite-temperature absorber must re-emit, and that re-emission is the
$J_{0}$ floor. Record silicon cells reach $V_{oc} \approx 0.74\,\si{
\volt}$; the gap between $0.74$ and $0.99\,\si{\volt}$ is non-radiative
recombination at defects, surfaces, and contacts, the dominant
real-cell voltage loss.

\begin{mastery}
The $J_{0}$ floor and the $0.13\,\si{\volt}$ gap between $V_{oc}$ and
the band gap come from detailed balance, a statement of the second law
applied to the cell. A cell at temperature $T_{c}$ in the dark emits
thermal photons above its gap at the rate a blackbody would, set by the
Bose-Einstein occupation of the photon modes. Under bias $V$ the
emitted flux rises by the factor $\exp(eV/k_{B}T_{c})$, because the
quasi-Fermi-level splitting that the bias imposes equals $eV$ and the
emitted photons carry a chemical potential $\mu = eV$. Setting absorbed
solar flux equal to emitted flux at open circuit gives $V_{oc}$
directly, with no reference to any material defect: the radiative limit
is the best any cell of that gap can do at that temperature and
illumination. The Boltzmann form $\exp(eV/k_{B}T)$ used in the diode law
is the high-energy tail of the exact Fermi-Dirac occupation
$f(E) = 1/[1+\exp((E-E_{F})/k_{B}T)]$. For carriers more than a few
$k_{B}T$ above the Fermi level the $+1$ in the denominator is
negligible and Boltzmann is accurate to better than a percent, which is
why the diode and solar-cell algebra of this chapter uses it. The
Boltzmann form fails inside a degenerate semiconductor, where heavy
doping pushes the Fermi level into the band and the occupation
saturates at one; there the full Fermi-Dirac integral is required, and
the effective density of states $N_{c}$, $N_{v}$ that hides the
statistics in the intrinsic-carrier formula is itself a Fermi-Dirac
result evaluated in the non-degenerate limit. The engineering content
is that detailed balance sets the ceiling and Boltzmann statistics give
the working algebra up to the degenerate limit.
\end{mastery}

\subsection{Fill factor and efficiency}

The cell delivers power $P = JV$, maximised at an operating point below
$V_{oc}$ and below $J_{sc}$. The \engterm{fill factor} measures how
square the $J$-$V$ knee is,
\[
\mathrm{FF} = \frac{J_{mp}V_{mp}}{J_{sc}V_{oc}},
\]
the ratio of the maximum-power rectangle to the $J_{sc}V_{oc}$
rectangle. For silicon a normalised voltage $v_{oc} = eV_{oc}/k_{B}T
\approx 29$ gives an ideal fill factor near $0.84$; series resistance
and shunt leakage pull it down to $0.80$ to $0.83$ in good cells. The
efficiency is
\[
\eta = \frac{J_{sc}\,V_{oc}\,\mathrm{FF}}{P_{\text{in}}}.
\]
For the record cell, $\eta = (42\,\si{\milli\ampere\per\centi\meter
\squared})(0.74\,\si{\volt})(0.83)/(100\,\si{\milli\watt\per\centi
\meter\squared}) = 0.258$, about $26\,\%$, matching the laboratory
record (current as of 2024).

\subsection{Real-cell losses and the module}

Three loss channels separate the ideal from the fielded panel.
Reflection at the bare silicon surface (refractive index $\approx 3.9$)
returns over a third of the incident light; a textured surface and a
$\sim 75\,\si{\nano\meter}$ silicon-nitride anti-reflection layer cut
this to a few percent. Series resistance in the emitter, the contacts,
and the metal grid burns power as $I^{2}R$ and is the main fill-factor
loss; the grid design trades shading area against resistance.
Recombination at surfaces, grain boundaries, and the rear contact
lowers both $V_{oc}$ and the collected current; passivated contacts and
a back-surface field are the modern remedies. Assembling cells into a
module adds further losses: the gaps between cells, the glass
reflection, and the resistive interconnect drop the panel efficiency to
about $21\,\%$ even when the cells are $24\,\%$, the last two bars of
Figure~\ref{fig:vol04ch12:sc-cell-loss-waterfall}. The IEA market
snapshot tracks the commercial fleet against these numbers
\cite{data:ch09t3-iea-pv-2023}.

\subsection{Verdict}

The silicon cell passes its own physics with little reserve. The
detailed-balance ceiling for the $1.12\,\si{\electronvolt}$ gap is
about $33\,\%$; the record laboratory cell reaches $26\,\%$, and a
commercial module about $21\,\%$. The dominant losses are, in order,
the spectral mismatch of a single gap to a broad spectrum (fixed only
by going to multiple junctions), non-radiative recombination (the
voltage gap from $0.99$ to $0.74\,\si{\volt}$), and the resistive and
optical losses of the metallisation and the module. The honest reading
is that silicon is within a few absolute percent of its single-junction
ceiling, so the remaining gains are incremental, and the large step to
$47\,\%$ requires the multi-junction stack of
Figure~\ref{fig:vol04ch12:shockley-queisser} with its added cost. The
case study has carried one cell from the solar spectrum to a signed
efficiency; the discipline, the chain from spectrum to band gap to
current to voltage to fill factor to module, is the transferable
result.

\section{An end-to-end case study: a white-LED luminaire}\pathtag{standard}

The solar cell ran the junction in reverse, turning photons into
current. The light-emitting diode runs the same junction forward,
turning current into photons, and a complete luminaire stacks a chain
of conversions between the wall socket and the lumens a room receives.
We take a phosphor-converted white-LED downlight and carry it from the
electrical input through the blue-emitter physics, the droop that limits
how hard the die may be driven, the phosphor that turns part of the blue
into yellow, the luminous efficacy that weights the emitted spectrum by
the eye, and the thermal derating that links junction temperature to
both efficiency and lifetime, ending in a luminaire an engineer would
specify. The device is a good second vehicle because every term in the
LED-laser-photovoltaic section reappears with the sign reversed, and
because the figure of merit, lumens per watt, is set not by a single
physical limit but by a product of six conversion efficiencies, each
with its own remedy.

\subsection{The target and the radiometric ceiling}

The luminaire is to deliver $1200\,\si{\lumen}$, the output of a retired
$75\,\si{\watt}$ incandescent bulb, at a correlated colour temperature
near $3000\,\si{\kelvin}$ (warm white) and a colour-rendering index above
$80$. The hard ceiling is set by the eye. Luminous flux is radiant power
weighted by the photopic luminosity function, which peaks at
$555\,\si{\nano\meter}$ where one watt of radiation is defined as
$683\,\si{\lumen}$. A source emitting all its power at that single green
wavelength would reach $683\,\si{\lumen\per\watt}$ of radiation and no
higher; any white source spreads power across wavelengths the eye weights
below unity, so a warm-white spectrum with good colour rendering has a
luminous efficacy of radiation near $300$ to $350\,\si{\lumen\per\watt}$
before a single electrical or optical loss is counted. The chain from
that radiometric ceiling down to the wall-plug number is the subject of
the case study, tracked in the waterfall of
Figure~\ref{fig:vol04ch12:led-efficiency-waterfall}.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-led-efficiency-waterfall.tex}

\subsection{The blue emitter and the wall-plug chain}

The emitter is an InGaN die emitting near $450\,\si{\nano\meter}$, the
material that won blue and white solid-state lighting. Its photon energy
by the working rule is $E = 1.24/0.45 = 2.76\,\si{\electronvolt}$, which
sets a radiative forward voltage near $2.76\,\si{\volt}$; a real die at
its operating current runs at about $V_{f} = 3.25\,\si{\volt}$, the
excess being series resistance and the junction's own ideality. The
electrical efficiency, the fraction of input power that reaches the
junction as useful drive rather than resistive heat, is the ratio of the
radiative voltage to the applied voltage, here about
$2.76/3.25 \approx 0.85$.

Inside the junction the internal quantum efficiency, the fraction of
injected electron-hole pairs that recombine radiatively rather than at
defects, reaches about $0.82$ for a good InGaN die at moderate current.
The light then has to escape a semiconductor of refractive index near
$2.4$; total internal reflection traps most of it, and even with a
shaped die, a dome encapsulant, and a roughened surface the extraction
efficiency is about $0.70$. The product so far,
\[
\eta_{\text{blue,WPE}}
= (0.85)(0.82)(0.70) \approx 0.49,
\]
is the wall-plug efficiency of the bare blue emitter: about half the
input electrical power leaves the die as blue photons. This is the first
three bars of Figure~\ref{fig:vol04ch12:led-efficiency-waterfall}.

\subsection{Droop and the operating point}

The internal quantum efficiency is not a constant; it peaks at low
current density and falls as the die is driven harder
(Figure~\ref{fig:vol04ch12:led-droop}). The droop is attributed mainly
to Auger recombination, a three-carrier process whose rate grows with
the cube of carrier density and so steepens as the current rises, with
carrier leakage out of the quantum wells contributing. A die that peaks
near $86\,\%$ at $10\,\si{\ampere\per\centi\meter\squared}$ falls to about
$77\,\%$ at the $35\,\si{\ampere\per\centi\meter\squared}$ a cost-driven
luminaire actually uses, because operating at the efficiency peak would
demand three to four times the die area for the same flux and the die is
the expensive part.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-led-droop.tex}

The design choice is explicit: each factor of two in drive current
roughly doubles the flux per die and so halves the die cost per lumen,
while costing a few points of efficiency to droop and adding heat that
must be removed. The luminaire designer reads the operating point off
this curve against a die-cost and a heat-sink budget, the trade that
sets how many dice the fixture carries.

\subsection{Phosphor down-conversion}

A blue die alone gives blue light. White light comes from coating the die
with a cerium-doped yttrium-aluminium-garnet (YAG:Ce) phosphor that
absorbs part of the blue and re-emits a broad yellow band; the unconverted
blue and the converted yellow mix to white. Two losses ride on the
conversion. The Stokes loss is the energy difference between an absorbed
blue photon ($2.76\,\si{\electronvolt}$) and an emitted yellow photon near
$580\,\si{\nano\meter}$ ($2.14\,\si{\electronvolt}$); the difference,
about $22\,\%$ of the converted photon's energy, becomes heat in the
phosphor. The phosphor's own quantum yield, the fraction of absorbed
photons it re-emits, is about $0.90$ for good YAG:Ce. For a warm-white
mix that converts most of the blue, the combined phosphor efficiency is
about $0.80$, dropping the running fraction from $49\,\%$ to about
$39\,\%$ of input power, the fifth bar of
Figure~\ref{fig:vol04ch12:led-efficiency-waterfall}. The phosphor also
sets the colour temperature and the colour-rendering index: more phosphor
and a red-emitting co-dopant warm the colour and lift the rendering index
above $90$, at a few more points of efficiency.

\subsection{Luminous efficacy and the lumen count}

The emitted spectrum, blue peak plus broad yellow, is now weighted by the
eye. The warm-white mix has a luminous efficacy of radiation near
$320\,\si{\lumen\per\watt}$, about $47\,\%$ of the $683\,\si{\lumen\per
\watt}$ monochromatic ceiling, because the spectrum puts power into the
blue and the deep red where the eye is insensitive. Folding this into the
running power fraction is bookkeeping: the luminous efficacy of the
luminaire is the product of the wall-plug radiometric efficiency and the
luminous efficacy of radiation. With $39\,\%$ of the input power leaving
as warm-white radiation at $320\,\si{\lumen\per\watt}$ of that radiation,
\[
\text{efficacy} = 0.39 \times 320\,\si{\lumen\per\watt}
\approx 125\,\si{\lumen\per\watt}
\]
at the die's rated temperature. A modern mid-power luminaire reaches
$120$ to $160\,\si{\lumen\per\watt}$ at the fixture level, so the estimate
lands in the right band. To deliver the target $1200\,\si{\lumen}$ at
$125\,\si{\lumen\per\watt}$ the luminaire draws
$1200/125 \approx 9.6\,\si{\watt}$, against the $75\,\si{\watt}$ of the
incandescent it replaces: the same light for an eighth of the power, the
number that retired the incandescent bulb.

\subsection{Thermal derating and lifetime}

The last bar of Figure~\ref{fig:vol04ch12:led-efficiency-waterfall} is
thermal. The die's efficiency and its emission wavelength both drift with
junction temperature, and the phosphor's quantum yield falls as it heats.
A die specified at $25\,{}^{\circ}\mathrm{C}$ but running at a junction
temperature of $85\,{}^{\circ}\mathrm{C}$ loses roughly $10$ to $15\,\%$
of its light output, the derating that pulls the running fraction from
about $28\,\%$ to about $24\,\%$ of input power and the efficacy from
$125$ down to a fielded $105$ to $110\,\si{\lumen\per\watt}$. The thermal
path is therefore a first-class design variable, not an afterthought: the
$9.6\,\si{\watt}$ drawn is almost all converted to heat at the die, since
only a quarter leaves as light, so a heat sink must carry about
$7\,\si{\watt}$ from a die a few millimetres across to hold the junction
near $85\,{}^{\circ}\mathrm{C}$.

Lifetime rides on the same temperature. LED light output does not fail
abruptly; it fades, and the industry rates lifetime as the time to reach
$70\,\%$ of initial output, the L70 figure. At a junction temperature near
$85\,{}^{\circ}\mathrm{C}$ a good luminaire reaches L70 at $30\,000$ to
$50\,000$ hours; every $10\,\si{\kelvin}$ of additional junction
temperature roughly halves that life, the same Arrhenius acceleration that
governs the flash-memory retention of the transistor section. The
dominant fade mechanisms are phosphor and encapsulant browning and
silver-reflector corrosion, all accelerated by heat, so the heat-sink
choice that holds efficiency also buys lifetime.

\subsection{Verdict}

The luminaire passes its specification with the heat sink as the binding
constraint. The chain from electrical input to delivered lumens runs
$0.85$ (electrical) $\times\,0.82$ (internal quantum) $\times\,0.70$
(extraction) $\times\,0.80$ (phosphor) for a wall-plug radiometric
efficiency near $39\,\%$, times a luminous efficacy of radiation near
$320\,\si{\lumen\per\watt}$, derated by temperature, for a fielded
efficacy near $105$ to $110\,\si{\lumen\per\watt}$ and a draw under
$10\,\si{\watt}$ for $1200\,\si{\lumen}$. The dominant losses are, in
order, the extraction loss at the high-index die surface, the phosphor
conversion loss, and the spectral mismatch of warm white to the eye;
droop and thermal derating are the levers the designer actually pulls,
because they trade against die cost and heat-sink size. The honest reading
is that the blue emitter is already near its practical efficiency, so the
remaining luminaire gains come from better extraction, narrower-band red
phosphors that waste less light in the deep red, and cooler operation; the
ceiling for warm white with good colour rendering sits near
$250\,\si{\lumen\per\watt}$, against which a fielded fixture at
$110\,\si{\lumen\per\watt}$ still has room. The case study has carried one
fixture from the wall socket to a delivered lumen count and a lifetime;
the discipline, the chain of six multiplied efficiencies each with its own
remedy, is the transferable result, and it is the photovoltaic waterfall
run backward.

\begin{mastery}
The luminous efficacy of radiation that caps the white-LED chain is a
weighting integral, not a single number, and the trade it encodes between
efficiency and colour quality is worth stating exactly. The luminous flux
in lumens is $\Phi_{v} = 683\,\si{\lumen\per\watt}\int V(\lambda)\,
\Phi_{e}(\lambda)\,d\lambda$, where $\Phi_{e}(\lambda)$ is the emitted
spectral power and $V(\lambda)$ is the photopic luminosity function,
unity at $555\,\si{\nano\meter}$ and falling to a few percent by
$450\,\si{\nano\meter}$ and $650\,\si{\nano\meter}$. A spectrum
concentrated at $555\,\si{\nano\meter}$ maximises the integral but renders
colour terribly, because objects reflecting red or blue have no light to
reflect; the colour-rendering index measures exactly that failure against
a reference illuminant. Warm white with a high rendering index therefore
spends power in the blue and red where $V(\lambda)$ is small, trading
luminous efficacy of radiation for fidelity. The design frontier is a
spectrum that places its power where the eye is sensitive and where common
surfaces reflect, which is why narrow-band red phosphors and quantum-dot
down-converters, with emission lines tuned to fall inside the photopic
band rather than spill into the deep red, are the route to luminaires that
hold a rendering index above $90$ without paying the usual efficacy
penalty. The same integral, weighted by the scotopic curve instead, runs
the argument for street lighting, where the dark-adapted eye's blue-shifted
sensitivity rewards a cooler spectrum.
\end{mastery}

\section{An end-to-end case study: a Fabry-Perot laser diode}\pathtag{standard}

The luminaire ran the junction forward to make incoherent light spread
across a broad spectrum. A laser diode runs the same forward junction
but adds an optical cavity, and the cavity changes the device from a
lamp into a source of coherent, near-monochromatic light with a sharp
turn-on at a threshold current. We take an edge-emitting Fabry-Perot
laser diode at $\lambda = 980\,\si{\nano\meter}$, the pump wavelength of
erbium-doped fibre amplifiers, and carry it from the active-region gain
through the threshold condition, the slope efficiency, the longitudinal
mode spacing, the beam divergence, and the temperature sensitivity,
ending in a P-I curve and a specification an engineer would sign. The
device is the natural third vehicle because it brings in the physics the
LED and the solar cell never needed: feedback, threshold, and
coherence, the three properties that separate a laser from a lamp. The
photonics here follow Saleh and Teich
\cite{text:saleh-teich-photonics}; the semiconductor numbers follow Sze
and Ng \cite{text:sze-semiconductor}.

\subsection{The active region and the gain}

The emitter is an InGaAs quantum well in a GaAs waveguide, a stripe
$L = 300\,\si{\micro\meter}$ long, $w = 3\,\si{\micro\meter}$ wide, with
an active-layer thickness $d = 8\,\si{\nano\meter}$. Forward bias
injects electrons and holes into the well; population inversion makes
the medium amplify rather than absorb light at the band-edge energy
$E = 1.24/0.98 = 1.27\,\si{\electronvolt}$. The modal gain rises with
injected current density from an absorbing medium, through a
transparency point where injection just balances absorption, into
amplification (Figure~\ref{fig:vol04ch12:laser-gain-loss}). For a
quantum-well laser the gain is approximately logarithmic in current,
$g(J) \approx g_0\ln(J/J_{tr})$, with a transparency current density
$J_{tr} \approx 50\,\si{\ampere\per\centi\meter\squared}$ and a gain
coefficient $g_0 \approx 12\,\si{\per\centi\meter}$ for this geometry.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-laser-gain-loss.tex}

\subsection{The threshold condition}

Light makes a round trip of the cavity, reflecting off the two cleaved
facets. Each facet of a bare GaAs-air interface reflects a fraction
$R = \big((n-1)/(n+1)\big)^{2} = (2.4/4.4)^{2} \approx 0.30$ of the
power, the rest leaking out as the useful beam. Oscillation is
self-sustaining when one round trip returns the light to its starting
amplitude, the round-trip gain equal to the round-trip loss,
\[
g_{th} = \alpha_i + \alpha_m,
\qquad
\alpha_m = \frac{1}{2L}\ln\!\left(\frac{1}{R_1 R_2}\right),
\]
where $\alpha_i$ is the distributed internal loss (free-carrier
absorption and scattering, here $\alpha_i \approx 10\,\si{\per\centi
\meter}$) and $\alpha_m$ is the mirror loss that accounts for the power
leaving through the facets. With $R_1 = R_2 = 0.30$ and $L = 300\,\si{
\micro\meter} = 0.03\,\si{\centi\meter}$,
\[
\alpha_m = \frac{1}{2(0.03)}\ln\!\left(\frac{1}{0.30\times 0.30}\right)
         = \frac{1}{0.06}\ln(11.1)
         = 16.7 \times 2.41
         = 40\,\si{\per\centi\meter},
\]
so the threshold gain is $g_{th} = 10 + 40 = 50\,\si{\per\centi\meter}$.
Inverting the logarithmic gain law for the current density that delivers
this gain, $J_{th} = J_{tr}\exp(g_{th}/g_0) = 50\exp(50/12) = 50\times
64 \approx 3.2\times 10^{3}\,\si{\ampere\per\centi\meter\squared}$.
Multiplying by the stripe area $wL = (3\times 10^{-4})(0.03) =
9\times 10^{-6}\,\si{\centi\meter\squared}$ gives a threshold current
\[
I_{th} = J_{th}\,wL \approx (3.2\times 10^{3})(9\times 10^{-6})
       \approx 29\,\si{\milli\ampere},
\]
the order of a real $980\,\si{\nano\meter}$ pump laser. The two levers
that lower threshold are visible in the algebra: a high-reflectivity
coating on the rear facet drops $\alpha_m$, and a shorter cavity raises
$\alpha_m$ but shrinks the area, so the cavity length is optimised
against the desired output power, not minimised.

\subsection{Slope efficiency and the P-I curve}

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-laser-pi-curve.tex}

Below threshold the device emits weak spontaneous light and behaves like
a dim LED; above threshold every additional injected electron pair feeds
stimulated emission and the output power climbs almost linearly with
current (Figure~\ref{fig:vol04ch12:laser-pi-curve}). The slope is the
\engterm{differential quantum efficiency}, the fraction of above-threshold
injected carriers that leave as photons through the output facet,
\[
\eta_d = \eta_i\,\frac{\alpha_m}{\alpha_i + \alpha_m},
\]
the internal quantum efficiency $\eta_i \approx 0.8$ scaled by the
fraction of the loss that is useful output coupling rather than internal
absorption. Here $\eta_d = 0.8\times 40/50 = 0.64$. Converting carriers
to optical power, the slope of the P-I curve is
\[
\frac{dP}{dI} = \eta_d\,\frac{h\nu}{e}
              = 0.64\times\frac{1.27\,\si{\electronvolt}}{e}
              = 0.64\times 1.27\,\si{\volt}
              \approx 0.81\,\si{\watt\per\ampere}
              = 0.81\,\si{\milli\watt\per\milli\ampere},
\]
since dividing the photon energy in electronvolts by the electron charge
returns a voltage numerically equal to the energy in volts. At a drive
current of $50\,\si{\milli\ampere}$, about $21\,\si{\milli\ampere}$ above
threshold, the output is $P = \eta_d (h\nu/e)(I - I_{th}) = 0.81\times
21 \approx 17\,\si{\milli\watt}$, a typical pump-laser rating. The
\engterm{wall-plug efficiency} folds in the forward voltage drop: at
$50\,\si{\milli\ampere}$ and a forward voltage near $1.5\,\si{\volt}$
the electrical input is $75\,\si{\milli\watt}$, so the wall-plug
efficiency is $17/75 \approx 0.23$, below the slope efficiency because
the threshold current is dissipated without producing useful light and
the forward voltage exceeds the photon voltage.

\subsection{Mode spacing and the spectrum}

The cavity admits only the longitudinal modes whose round-trip phase is
a multiple of $2\pi$, standing waves with an integer number of
half-wavelengths between the facets. Adjacent modes differ by one
half-wavelength, and their wavelength spacing is
\[
\Delta\lambda = \frac{\lambda^{2}}{2 n_g L},
\]
with $n_g \approx 3.6$ the group index of the GaAs waveguide. For
$\lambda = 980\,\si{\nano\meter}$ and $L = 300\,\si{\micro\meter}$,
\[
\Delta\lambda = \frac{(980\times 10^{-9})^{2}}
                     {2(3.6)(300\times 10^{-6})}
             = \frac{9.6\times 10^{-13}}{2.16\times 10^{-3}}
             = 4.4\times 10^{-10}\,\si{\meter}
             = 0.44\,\si{\nano\meter}.
\]
A bare Fabry-Perot laser oscillates on several of these modes at once,
spaced $0.44\,\si{\nano\meter}$ apart under the gain envelope, so its
spectrum is a comb a few modes wide. A single-frequency source for
telecom requires a wavelength-selective element, a distributed-feedback
grating built into the waveguide, which suppresses all but one mode and
narrows the linewidth below $10\,\si{\mega\hertz}$, the structure noted
in the LED-laser-photovoltaic section.

\subsection{Beam divergence}

The light leaving the facet escapes from an aperture only a wavelength or
two across, so diffraction spreads it into a wide, elliptical beam. The
far-field half-angle is set by the diffraction relation $\theta \approx
\lambda/(\pi w_{0})$ for a near-field spot of size $w_{0}$. The fast axis,
perpendicular to the thin $8\,\si{\nano\meter}$ active layer, has the
smallest aperture and so the largest divergence, of order $30^{\circ}$
half-angle; the slow axis, set by the $3\,\si{\micro\meter}$ stripe
width, diverges only about $8^{\circ}$. The output is therefore a
strongly astigmatic elliptical beam, which is why a laser-diode module
carries a fast-axis collimating lens and often an anamorphic prism pair
to circularise the beam before it enters a fibre. The divergence is not a
defect to be polished out; it is the diffraction limit of a
wavelength-scale aperture, and the cure is collimation, not a better
facet.

\subsection{Temperature sensitivity}

The threshold current rises with temperature because the carriers spread
over a wider energy range and more of them leak out of the well, an
empirical exponential,
\[
I_{th}(T) = I_{th}(T_0)\,\exp\!\big((T - T_0)/T_0^{*}\big),
\]
with a characteristic temperature $T_0^{*} \approx 120\,\si{\kelvin}$
for a good InGaAs device (lower, near $50$ to $70\,\si{\kelvin}$, for
the InP lasers at $1.55\,\si{\micro\meter}$). A rise from $25$ to
$60\,{}^{\circ}\mathrm{C}$, a span of $35\,\si{\kelvin}$, multiplies the
threshold by $\exp(35/120) = \exp(0.29) = 1.34$, so a $29\,\si{\milli
\ampere}$ threshold climbs to about $39\,\si{\milli\ampere}$, the shift
between the two laser curves of
Figure~\ref{fig:vol04ch12:laser-pi-curve}. At a fixed drive current the
output power then drops sharply, and the emission wavelength itself
drifts about $0.3\,\si{\nano\meter\per\kelvin}$ as the gap shrinks with
temperature. A laser transmitter therefore carries either a
thermoelectric cooler holding the die at a set point or a monitor
photodiode behind the rear facet feeding an automatic-power-control loop
that raises the current to hold output constant as the device heats.

\subsection{Verdict}

The laser diode passes its specification with the threshold and its
temperature drift as the binding constraints. The chain runs from the
logarithmic gain law through the threshold condition $g_{th} = \alpha_i +
\alpha_m \approx 50\,\si{\per\centi\meter}$, giving $I_{th} \approx
29\,\si{\milli\ampere}$; through the differential efficiency $\eta_d =
0.64$ and a slope near $0.8\,\si{\milli\watt\per\milli\ampere}$, giving
$17\,\si{\milli\watt}$ at $50\,\si{\milli\ampere}$; through a longitudinal
mode spacing of $0.44\,\si{\nano\meter}$ that makes the bare device
multimode; through a strongly astigmatic beam needing collimation; to a
threshold that climbs by a third over a $35\,\si{\kelvin}$ heating. The
dominant design levers are the facet reflectivities, which set both the
threshold and the slope through their opposite pulls on $\alpha_m$, and
the thermal management, which sets the usable current range. The honest
reading is that the cavity does the work the LED never asked of it,
turning a spontaneous lamp into a coherent source, and the price is the
threshold current burned without output and a device that must be held at
temperature. The case study has carried one diode from gain to a P-I
curve and a beam specification; the discipline, the chain from gain
through threshold and slope to mode, beam, and temperature, is the
transferable result, and it is the LED with feedback and a threshold
added.

\begin{mastery}
The threshold condition $g_{th} = \alpha_i + \alpha_m$ is the
steady-state of the rate equations that govern the laser's dynamics, and
those equations carry the device behaviour the static P-I curve hides.
Write $N$ for the carrier density in the well and $S$ for the photon
density in the lasing mode. Carriers are injected at rate $I/eV$, decay
spontaneously at rate $N/\tau_n$, and are consumed by stimulated emission
at rate $g(N)\,v_g\,S$; photons are created by that same stimulated
process, leave the cavity at rate $S/\tau_p$, and receive a small
spontaneous seed,
\[
\frac{dN}{dt} = \frac{I}{eV} - \frac{N}{\tau_n} - g(N)\,v_g\,S,
\qquad
\frac{dS}{dt} = \Gamma g(N)\,v_g\,S - \frac{S}{\tau_p}
              + \beta\frac{N}{\tau_n},
\]
with $v_g$ the group velocity, $\Gamma$ the confinement factor, $\tau_p$
the photon lifetime $1/(v_g(\alpha_i+\alpha_m))$, and $\beta$ the tiny
fraction of spontaneous emission that enters the lasing mode. Below
threshold the photon term cannot grow because the gain has not reached
the loss, so $S$ stays at the spontaneous seed and $N$ rises with current;
at threshold the bracket $\Gamma g v_g - 1/\tau_p$ passes through zero, $S$
grows, and the stimulated term clamps $N$ near its threshold value. That
clamping is why the P-I curve is linear above threshold: added current
goes almost entirely into photons because the carrier density, and hence
the gain, can no longer rise. Perturbing the steady state gives a damped
oscillation between $N$ and $S$, the relaxation-resonance frequency that
sets the laser's modulation bandwidth, typically a few gigahertz and
rising with the square root of the output power. The static slope
efficiency and the dynamic bandwidth both fall out of these two coupled
equations; the engineering content is that the same rate balance that
fixes the threshold also fixes how fast the laser can be modulated to
carry data.
\end{mastery}

\section{The transistor as a quantum device}\pathtag{standard}

The \engterm{metal-oxide-semiconductor field-effect transistor}
(MOSFET) is the standard device of digital electronics. A thin oxide
insulates a metal (or polysilicon) gate from a semiconductor channel
between two doped source and drain regions. Voltage on the gate
modulates the carrier density in the channel by capacitive coupling;
the source-to-drain current is the working output.

In the long-channel regime (channel length $\gg$ de~Broglie
wavelength), the MOSFET is well-described by the drift-diffusion
equations of classical semiconductor physics, with thresholds and
saturation behaviour captured by the standard Pao-Sah or Sah-Shockley
equations. In the short-channel regime, ballistic transport (the
electron crosses the channel without scattering), quantum tunnelling
through the oxide barrier, and quantum confinement of the channel
carriers become important. Modern $5\,\si{\nano\meter}$-process
nodes (TSMC N5, Samsung 5LPP, Intel 4) have channel lengths of $\sim
6$-$10\,\si{\nano\meter}$ and rely on quantum-mechanical simulations
during device design.

The \engterm{quantum tunnelling} through the gate oxide sets the
leakage current of every modern transistor. The tunnelling current
density through a barrier of height $\phi$ and thickness $d$ scales
as
\[
J_{\text{tunnel}} \sim \exp(-2d\sqrt{2m^{*}\phi}/\hbar).
\]
For SiO$_{2}$ with $\phi \approx 3.1\,\si{\electronvolt}$, doubling
$d$ reduces leakage by a factor of $\sim 10^{4}$; halving $d$
increases it by the same factor
(Figure~\ref{fig:vol04ch12:tunneling-well}). The script
\texttt{code/tunneling\_leakage.py} tabulates this exponential
sensitivity against oxide thickness.

\subsection{Where the tunnelling exponent comes from: the WKB
transmission coefficient}\pathtag{mastery}

The exponential in the leakage formula is not an empirical fit; it
falls out of the wave equation in three lines, and the derivation is
worth carrying because it shows exactly which device parameters live
in the exponent and which sit in the harmless prefactor. A particle of
energy $E$ meets a potential $V(x)$ that exceeds $E$ over a barrier
region. Inside the barrier the time-independent Schr\"odinger equation
\[
-\frac{\hbar^{2}}{2m^{*}}\frac{d^{2}\psi}{dx^{2}} + V(x)\psi = E\psi
\]
rearranges to $\psi'' = \kappa(x)^{2}\psi$ with a real, positive local
decay rate
\[
\kappa(x) = \frac{\sqrt{2m^{*}\big(V(x) - E\big)}}{\hbar}.
\]
Where $E > V$ the right-hand side would carry a minus sign and the
solution oscillates; where $E < V$ it grows or decays. The physical
solution inside a thick barrier is the decaying one. When the barrier
varies slowly on the scale of the local decay length $1/\kappa$, the
Wentzel-Kramers-Brillouin (WKB) approximation builds the wavefunction
by exponentiating the accumulated decay,
\[
\psi(x) \approx \psi(x_1)\,
\exp\!\left(-\int_{x_1}^{x}\kappa(x')\,dx'\right),
\]
the local-decay-rate generalisation of the constant-$\kappa$ exponential
$e^{-\kappa x}$. The transmission probability is the squared ratio of the
amplitude leaving the barrier at $x_2$ to the amplitude entering it at
$x_1$, which is the square of that exponential,
\[
T \approx \exp\!\left(-2\int_{x_1}^{x_2}
\frac{\sqrt{2m^{*}\big(V(x) - E\big)}}{\hbar}\,dx\right).
\]
The factor of two is the move from amplitude to probability. For a
rectangular barrier of constant height $\phi = V - E$ and width $d$ the
integral is trivial and the WKB result collapses to the leakage formula,
\[
T \approx \exp\!\left(-2d\,
\frac{\sqrt{2m^{*}\phi}}{\hbar}\right)
= \exp(-2\kappa d),
\qquad
\kappa = \frac{\sqrt{2m^{*}\phi}}{\hbar},
\]
plotted against thickness for two barrier heights in
Figure~\ref{fig:vol04ch12:wkb-transmission}. Three device readings come
straight off the exponent. The thickness $d$ enters linearly inside the
exponential, so leakage is exponential in oxide thickness, the origin of
the five-orders-of-magnitude swing in the worked example. The barrier
height $\phi$ enters under a square root, so a material with a higher
band offset to silicon tunnels less at the same physical thickness, one
reason a higher-offset dielectric helps even before the capacitance
argument. The effective mass $m^{*}$ enters the same way, which is why
the tunnelling mass in the insulator, not the silicon channel mass, is
the number that must be measured. The prefactor that WKB drops, of order
unity and weakly bias-dependent, changes the leakage by factors of a few;
the exponent changes it by factors of thousands, so a design estimate
that gets the exponent right and the prefactor to an order of magnitude
is the working tool. A triangular barrier under field, where $V(x)$
slopes because a voltage is dropped across the oxide, is the same integral
with a linear $V(x)$, and evaluating it returns the Fowler-Nordheim law
that governs flash-memory writing and erase; the field tilts the barrier,
shortens the tunnelling distance at fixed oxide thickness, and turns the
ten-year retention barrier into a microsecond write path.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-wkb-transmission.tex}

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-tunneling-well.tex} The scaling motivated the industry
transition from SiO$_{2}$ to high-$k$ dielectrics (HfO$_{2}$) around
2007, allowing the same effective capacitance with a thicker
physical layer and lower tunnelling.

A \engterm{flash memory} cell stores information in a floating gate
isolated from the channel by a thin oxide. Writing the cell tunnels
electrons through the oxide onto the floating gate; reading senses
the threshold shift the trapped charge produces; erasing reverses
the tunnelling. The cell's retention (the time over which the
trapped charge persists) is set by the same exponential tunnelling
formula and is $\sim 10$ years at room temperature with a typical
oxide thickness of $\sim 8\,\si{\nano\meter}$. The retention
degrades at elevated temperature and after many write-erase cycles
(the wear of the oxide); this is the working specification of every
solid-state storage device.

\subsection{The subthreshold swing and why the transistor cannot be
switched off}\pathtag{standard}

The tunnelling leakage is the gate-oxide's contribution to standby power,
but a second, deeper floor sits under every transistor and it is set not by
tunnelling but by the same Boltzmann statistics that governed the diode.
When the gate is biased below threshold the channel is not empty; it holds a
thermally populated tail of carriers whose density follows the Boltzmann
factor of the surface potential, so the drain current does not switch off at
threshold but falls exponentially as the gate voltage is lowered, the
\engterm{subthreshold} regime. The gate swing needed to drop the current by
one decade is the \engterm{subthreshold swing},
\[
S = \frac{dV_{GS}}{d(\log_{10} I_D)}
  = \ln(10)\,\frac{k_BT}{e}\left(1 + \frac{C_{dep}}{C_{ox}}\right),
\]
the thermal voltage times $\ln 10$, scaled by a body factor that measures
how well the gate controls the surface potential against the depletion
capacitance beneath it. The leading term $\ln(10)\,k_BT/e$ at room
temperature is $2.303\times 26\,\si{\milli\volt} = 60\,\si{\milli\volt}$ per
decade, and the body factor only raises it, so a real transistor needs
$70$ to $90\,\si{\milli\volt}$ of gate swing for each factor of ten in
current.

That $60\,\si{\milli\volt}$ per decade is a hard floor for any device whose
carriers obey Boltzmann statistics, and it sets the whole power budget of
digital logic. To switch a transistor from a leakage floor to a full on
current across a swing of, say, five decades demands at least $5\times 60 =
300\,\si{\milli\volt}$ of gate swing, and the supply voltage must exceed
that with margin. The swing cannot be lowered by better fabrication because
it is thermal: it scales with $T$, so a cooled circuit switches more sharply,
and a room-temperature circuit is stuck at the Boltzmann limit. The one
escape is a device whose carriers do not cross a thermal barrier at all,
which is why the tunnel field-effect transistor, injecting carriers by
band-to-band tunnelling rather than over a barrier, is pursued as the route
to a sub-$60\,\si{\milli\volt}$ swing and a lower-voltage logic. The
engineering content is that the same Boltzmann tail that gives the diode its
$60\,\si{\milli\volt}$-per-decade slope gives the transistor its
$60\,\si{\milli\volt}$-per-decade off-switch, and the supply voltage of every
processor is bounded below by it.

\begin{mastery}
The subthreshold floor and the tunnelling floor together draw the box the
transistor scaling must live in, and the tension between them is worth
stating exactly because it is why voltage scaling stalled. Lowering the
supply voltage saves dynamic power, which scales as $CV^{2}$, so every volt
is precious; but the on-off current ratio a logic gate needs is fixed, and
at a fixed subthreshold swing that ratio costs a fixed gate swing, so the
threshold voltage cannot be pushed down without either raising the off-state
leakage or steepening the swing. Steepening the swing means raising
$C_{ox}$, a thinner or higher-permittivity gate dielectric, which pushes the
oxide toward the tunnelling floor of the leakage example; so the two floors
squeeze from opposite sides, the Boltzmann swing setting how much voltage a
decade of current costs and the tunnelling limit setting how thin the oxide
that steepens the swing may be. The high-$k$ dielectric transition bought
one round of relief by raising $C_{ox}$ at a physical thickness that keeps
tunnelling in check, and the fin and gate-all-around geometries bought
another by wrapping the gate around the channel to drive $C_{dep}/C_{ox}$
toward zero and the swing toward its $60\,\si{\milli\volt}$ ideal. Both are
moves inside the same box, not escapes from it; the escape requires changing
the injection mechanism away from thermal emission, the tunnel-transistor
route. The engineering content is that two quantum-statistical floors, one
from the Boltzmann tail and one from barrier tunnelling, jointly fix the
lowest voltage at which a logic gate can be built from a barrier-controlled
channel.
\end{mastery}

\section{MRI, electron microscopy, X-ray}\pathtag{standard}

Three imaging modalities at the engineering quantum scale are MRI
(magnetic resonance imaging), electron microscopy, and X-ray
imaging.

\engterm{MRI} uses the nuclear magnetic resonance of hydrogen nuclei
(protons) in tissue. In a static magnetic field $B_{0}$, the proton's
two spin states are split in energy by $\Delta E = \gamma\hbar B_{0}$,
where $\gamma = 2.675\times 10^{8}\,\si{\radian\per\second\per\tesla}$
is the gyromagnetic ratio of the proton. At $B_{0} = 1.5\,\si{\tesla}$,
$\Delta E$ corresponds to a Larmor frequency of $63.9\,\si{\mega
\hertz}$; at $3\,\si{\tesla}$, $127.7\,\si{\mega\hertz}$. The standard
MRI sequence applies a $90^{\circ}$ RF pulse at the Larmor frequency
to flip the equilibrium magnetisation into the transverse plane,
measures the free-induction decay or echo signal, and reconstructs
the image from spatial-frequency information encoded by gradient
fields. The spatial resolution is $\sim 1\,\si{\milli\meter}$ at
clinical field strengths and below $100\,\si{\micro\meter}$ at high
field for small-bore imaging.

\engterm{Electron microscopy} uses an electron beam at $10$ to
$300\,\si{\kilo\volt}$ accelerating voltage as the imaging probe.
The de~Broglie wavelength of a $100\,\si{\kilo\volt}$ electron is
$\sim 3.7\,\si{\pico\meter}$, far smaller than visible-light
wavelengths and capable in principle of resolving individual atoms.
Practical resolution is limited by lens aberrations (electron
optics suffers from spherical aberration that is much harder to
correct than in light optics) to $\sim 0.1\,\si{\nano\meter}$ with
aberration-corrected instruments. The transmission electron
microscope (TEM) images electrons that traverse a thin sample; the
scanning electron microscope (SEM) rasters a focused beam and detects
secondary electrons, giving surface-topology images of bulk samples.

\engterm{X-ray imaging} uses photons of energy $1$ to $200\,\si{
\kilo\electronvolt}$, with $\lambda$ from $0.01$ to $1\,\si{\nano
\meter}$, generated by accelerated electrons hitting a metal target
(bremsstrahlung and characteristic emission) or by synchrotron
radiation from relativistic electrons in a storage ring. Medical
X-ray imaging exploits differential absorption between bone and soft
tissue (bone's higher electron density absorbs more); X-ray
diffraction maps the crystal structure of materials; X-ray
photoelectron spectroscopy maps surface composition. The quantum
content is that the X-ray is a photon whose interaction probability
is set by the photoelectric absorption cross section, scaling roughly
as $Z^{4}/E^{3}$ in the diagnostic range.

The three modalities share a quantum logic worth stating together. Each
probes matter with a quantum whose energy or wavelength is chosen to
match the structure being resolved, and each is limited by the
statistics of that quantum. MRI works at radio-frequency energies far
below $k_{B}T$, so the spin population difference between the two proton
states is tiny, of order one part in $10^{5}$ at $1.5\,\si{\tesla}$ and
body temperature; the signal comes from the small net magnetisation of
an enormous number of protons, and the drive to higher field is a drive
for more signal from a larger population difference. Electron microscopy
works at the opposite extreme, picometre wavelengths that in principle
resolve atoms, where the limit is not the wavelength but the lens
aberration and the radiation dose the specimen tolerates before it is
destroyed. X-ray imaging sits between, with wavelengths near atomic
spacings and a contrast set by the $Z^{4}/E^{3}$ absorption that makes
bone visible against tissue, and a dose budget that trades image quality
against the ionisation the patient absorbs. In every case the engineer
chooses the quantum to the job and then lives with its shot-noise floor:
the number of detected quanta, not the optics, sets the achievable
contrast at the limit.

\subsection{The MRI signal and why field strength buys it}\pathtag{mastery}

The claim that MRI works from a population difference of one part in
$10^{5}$ is worth carrying to a number, because it explains the field-strength
arms race and the whole cost structure of the machine. The two proton spin
states in a field $B_0$ are split by $\Delta E = \gamma\hbar B_0$, and at
thermal equilibrium their populations follow the Boltzmann ratio. The
fractional population difference that produces the net magnetisation is
\[
\frac{\Delta N}{N} = \tanh\!\left(\frac{\gamma\hbar B_0}{2k_B T}\right)
\approx \frac{\gamma\hbar B_0}{2k_B T},
\]
the high-temperature limit because the splitting is minute against $k_B T$.
At $B_0 = 1.5\,\si{\tesla}$ and body temperature, $\gamma\hbar B_0 =
(2.675\times 10^{8})(1.055\times 10^{-34})(1.5) = 4.2\times 10^{-26}\,\si{
\joule}$ against $2k_B T = 2(1.38\times 10^{-23})(310) = 8.6\times 10^{-21}
\,\si{\joule}$, a ratio of $4.9\times 10^{-6}$, about five protons per million
in excess. The signal is small only because the excess is small; it is
usable only because a cubic millimetre of water holds of order $10^{20}$
protons, so the net moment comes from $\sim 10^{15}$ excess spins, an
enormous absolute number. Two scalings follow, and both drive the machine
toward higher field. The equilibrium magnetisation is proportional to $B_0$
through the population difference, and the voltage the precessing
magnetisation induces in the receive coil is proportional to the Larmor
frequency, itself proportional to $B_0$; the signal power therefore climbs
faster than linearly in field while the dominant body noise climbs roughly
linearly, so the signal-to-noise ratio rises with field. That is the entire
argument for $3\,\si{\tesla}$ clinical scanners and $7\,\si{\tesla}$ research
machines, and the cost is the superconducting magnet, its cryogenics, and the
tighter demands on field uniformity that a higher frequency imposes. The
resolution follows from the same signal budget rather than from any
wavelength: the RF wavelength at $64\,\si{\mega\hertz}$ is metres, far larger
than the body, so MRI resolution is set not by diffraction but by how finely
the gradient fields can encode position before the voxel holds too few excess
spins to rise above the noise, which is why sub-millimetre imaging needs the
signal that only high field and long averaging supply.

\subsection{The resolution limit of the electron microscope}\pathtag{mastery}

The electron microscope reaches picometre wavelengths yet resolves only
tenths of a nanometre, a thousandfold gap between the wavelength and the
achieved resolution that the diffraction limit alone does not explain, and
the reason is worth stating because it inverts the usual optical intuition.
A round magnetic electron lens cannot be made converging for all rays: it
always focuses off-axis rays too strongly, an unavoidable positive spherical
aberration that has no simple corrector of the kind a glass lens accepts.
The blur it produces grows with the cube of the aperture half-angle $\alpha$,
$d_s = C_s\alpha^{3}$, while the diffraction blur falls with aperture as
$d_d = 0.61\lambda/\alpha$; the two pull opposite ways in $\alpha$, so there
is an optimum aperture that minimises their sum. Balancing them, the best
resolution scales as
\[
d_{\min} \approx A\,(C_s\lambda^{3})^{1/4},
\]
with $A$ of order unity. For a $200\,\si{\kilo\volt}$ instrument
($\lambda \approx 2.5\,\si{\pico\meter}$) and a spherical-aberration
coefficient $C_s \approx 1\,\si{\milli\meter}$, the fourth root of $C_s
\lambda^{3} = (10^{-3})(2.5\times 10^{-12})^{3} = 1.6\times 10^{-38}\,
\text{m}^{4}$ is about $3.5\times 10^{-10}\,\si{\meter}$, a resolution
near $0.35\,\si{\nano\meter}$ despite the picometre wavelength. The resolution
is limited by the lens, not the wavelength, and the payoff of that formula is
that resolution improves only as the fourth root of $C_s$: halving the
aberration buys just a $16\,\%$ improvement, so brute-force lens polishing
stalls. The decisive advance was the aberration corrector, a multipole
element that generates the negative spherical aberration a round lens cannot,
cancelling $C_s$ and pushing resolution below $0.1\,\si{\nano\meter}$ into the
regime where individual atomic columns are resolved. A second limit sits
underneath: the specimen tolerates only a finite electron dose before
radiation damage destroys the structure being imaged, so the achievable
resolution on a beam-sensitive sample, a protein or a battery interface, is
set by how many electrons can be spent before the sample is gone, the same
dose-against-information trade that governs the X-ray of a patient.

\subsection{The X-ray tube spectrum and the Duane-Hunt limit}\pathtag{mastery}

The X-ray section named bremsstrahlung and characteristic emission as the two
sources in a tube without saying what fixes the spectrum, and the shape is worth
carrying because it is a photoelectric effect run backward and it sets the tube
voltage a diagnostic machine must supply. An electron accelerated through a tube
voltage $V$ arrives at the target carrying kinetic energy $eV$, and when it is
deflected by a target nucleus it radiates a photon, the bremsstrahlung. The
photon energy is whatever fraction of the electron's energy the single
deceleration event converts, so most photons are soft and the spectrum rises
toward low energy; but no photon can carry more energy than the electron brought,
which places a sharp ceiling at the short-wavelength end,
\[
h\nu_{\max} = eV,
\qquad
\lambda_{\min} = \frac{hc}{eV}
              = \frac{1.24\,\si{\nano\meter}}{V[\si{\kilo\volt}]},
\]
the Duane-Hunt limit. The rule is the photoelectric relation with the roles of
photon and electron exchanged: there, a photon of energy $h\nu$ gave up its
energy to an electron; here, an electron of energy $eV$ gives up at most all of
it to one photon. For a diagnostic tube at $120\,\si{\kilo\volt}$ the shortest
wavelength is $\lambda_{\min} = 1.24/120 = 0.0103\,\si{\nano\meter}$, a
$120\,\si{\kilo\electronvolt}$ photon, and the whole bremsstrahlung continuum
lies at longer wavelengths below it. Riding on that smooth continuum are sharp
characteristic lines: an incident electron that knocks a $K$-shell electron out
of a target atom leaves a vacancy that an outer electron fills, emitting a photon
at the fixed energy difference of the two shells, the tungsten $K\alpha$ line
near $59\,\si{\kilo\electronvolt}$ for a tungsten anode. Those line energies
follow Moseley's law, the transition energy scaling as $(Z-1)^{2}$ with the
atomic number of the target, which is why a heavy element such as tungsten is
chosen for the anode: high $Z$ places the characteristic lines at diagnostic
energies and the dense nucleus makes the bremsstrahlung efficient. Two engineering
readings come off the spectrum. Raising the tube voltage both extends the
short-wavelength ceiling to harder, more penetrating photons and increases the
radiated efficiency, which climbs roughly as $ZV$, though even at diagnostic
voltages under one percent of the beam power leaves as X-rays and the rest heats
the anode, the reason a rotating water-cooled anode is standard. Filtering the
soft end of the continuum with a thin aluminium or copper sheet hardens the beam,
removing the low-energy photons that would be absorbed in the patient's skin
without reaching the detector, so the same spectrum that the tube voltage sets is
shaped by filtration to trade dose against contrast, the clinical version of the
dose-against-information budget the imaging section named.

\section{An applied vignette: the uncooled microbolometer}\pathtag{standard}

The three imaging modalities chose a quantum to match the structure being
resolved. A thermal-imaging camera makes the opposite choice: it gives up
the photon entirely and senses heat. The uncooled microbolometer is the
device behind every low-cost thermal camera, and it is the chapter's
counterexample, a quantum-scale sensor whose figure of merit is set by
thermodynamic fluctuation rather than by photon statistics, worth a short
study because the contrast sharpens what the photodiode sections assumed.

A scene at room temperature radiates as a blackbody whose Planck peak,
by Wien's law at $T = 300\,\si{\kelvin}$, sits at $\lambda_{\max} =
2.898\times 10^{-3}/300 = 9.7\,\si{\micro\meter}$, in the long-wave
infrared. A photodiode that detects these photons by band-gap absorption
would need a gap below $hc/\lambda = 1.24/9.7 = 0.13\,\si{\electronvolt}$,
a narrow-gap material such as mercury-cadmium-telluride, and at room
temperature such a gap is swamped by its own thermal carriers: $n_i \sim
\exp(-E_g/2k_B T)$ with $E_g = 0.13\,\si{\electronvolt}$ and $k_B T =
0.026\,\si{\electronvolt}$ gives an exponent of only $-2.5$, so the dark
current buries the signal. Cooled photon detectors solve this by holding
the focal plane near $77\,\si{\kelvin}$, the route taken in
high-performance thermal imagers, at the cost of a cryocooler. The
microbolometer takes the other road
(Figure~\ref{fig:vol04ch12:microbolometer-pixel}): a thin absorbing
membrane carrying a temperature-sensitive resistor is suspended on legs of
high thermal resistance above the silicon readout circuit, so that
absorbed infrared power raises the membrane temperature and the readout
measures the resistance change. No band gap, no cooling.

The pixel is a thermal balance. Absorbed power $P$ flows out through the
support legs of thermal conductance $G$, so the steady temperature rise is
$\Delta T = P/G$, and the membrane's heat capacity $C$ sets the thermal
time constant $\tau = C/G$. A typical $17\,\si{\micro\meter}$ pixel has a
thermal conductance near $G = 1\times 10^{-7}\,\si{\watt\per\kelvin}$ and a
time constant of order $10\,\si{\milli\second}$, comfortably faster than a
video frame. The scene contrast that matters is the change in incident
power when a target in the field warms by one kelvin. Differentiating the
Stefan-Boltzmann law, the radiated power changes by $dP/dT = 4\sigma T^{3}$
per unit area, about $6\,\si{\watt\per\meter\squared\per\kelvin}$ at room
temperature; folded through the lens solid angle, the optical
transmission, and the pixel area, a one-kelvin scene change deposits on the
order of $10^{-10}\,\si{\watt}$ on the membrane, which the small $G$ turns
into a membrane rise of about a millikelvin. The thermistor, with a
temperature coefficient of resistance near $-2\,\%$ per kelvin for the
vanadium-oxide or amorphous-silicon films used, converts that millikelvin
into a resistance change of a few parts in $10^{5}$, the signal the readout
must pull out of its noise.

The performance figure is the \engterm{noise-equivalent temperature
difference} (NETD), the scene temperature change that produces a signal
equal to the noise, and it is set by thermal physics, not photon shot
noise. The deepest floor is the thermal-fluctuation noise of the membrane
itself: a body of heat capacity $C$ exchanging heat through a conductance
$G$ has mean-square temperature fluctuation $\langle\Delta T^{2}\rangle =
k_B T^{2}/C$ by equipartition of the thermal energy, which sets a floor on
how small a steady $\Delta T$ the pixel can resolve. Good uncooled arrays
reach an NETD of $30$ to $50\,\si{\milli\kelvin}$, enough to see a person
against a wall, while the cooled photon imager reaches a few millikelvin.
The microbolometer trades sensitivity for the elimination of the
cryocooler, and that trade is why thermal imaging moved from a military
instrument to a phone accessory and a building-inspection tool. The
engineering content is the inversion of the chapter's usual rule: where the
photodiode sections chose the quantum to the job and lived with its shot
noise, the microbolometer refuses the quantum, senses bulk heat, and lives
instead with the thermodynamic fluctuation of a microscopic membrane. The
quantum floor and the thermal floor are two faces of the same statistical
limit, and which one binds a sensor is the first question to ask of any
detector specification.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-microbolometer-pixel.tex}

\section{An applied vignette: the resonant-tunnelling diode}\pathtag{mastery}

The tunnelling section treated the barrier as a leak to be suppressed. A
resonant-tunnelling diode makes the leak the whole device, and it is worth a
short study because it turns quantum confinement and tunnelling, the two
effects the transistor section fights, into a working circuit element with a
property no classical device has: a current that falls as the voltage rises.

The structure is a thin quantum well, a few nanometres of a narrow-gap
material such as InGaAs, sandwiched between two even thinner barriers of a
wider-gap material such as AlAs, with heavily doped emitter and collector
outside. The well is narrow enough that its particle-in-a-box confinement,
the same $E_n = n^{2}\pi^{2}\hbar^{2}/(2m^{*}L^{2})$ ladder the quantum-well
laser used, gives it a single bound state well above the well bottom. An
electron can cross the double barrier only when its energy lines up with that
bound state; off resonance the two barriers in series block it almost
completely, but on resonance the well acts as a matched cavity and the
transmission through both barriers rises toward unity, the electron analogue
of a Fabry-Perot filter passing a wavelength that fits its cavity.

The current-voltage curve of Figure~\ref{fig:vol04ch12:rtd-iv} follows from
sweeping that alignment. At zero bias the well level sits above the emitter
Fermi sea and no resonant channel is open. Raising the bias pulls the well
level down through the filled emitter states, and the resonant current climbs
to a peak when the level aligns with the bulk of the emitter electrons. Push
the bias further and the well level drops below the bottom of the emitter
conduction band; there are no longer any emitter electrons at the resonant
energy, the channel closes, and the current falls even as the voltage keeps
rising. That falling stretch is the region of \engterm{negative differential
resistance}, where $dI/dV < 0$, and it is the property that makes the device
useful. Past the valley, ordinary non-resonant tunnelling and thermionic
emission over the barriers take over and the current climbs again, so the
curve is a peak, a valley, and a final rise. The ratio of the peak current to
the valley current is the device's figure of merit, a few to a few tens in
good room-temperature material.

A negative differential resistance cancels the positive resistance of a
resonant tank, so an RTD biased in its falling region and placed across an
inductor and capacitor sustains an oscillation the way an amplifier with
feedback would, without a transistor. Because the transit through the
few-nanometre stack is faster than any charge-storage delay, RTD oscillators
reach into the terahertz band that ordinary transistors struggle to attain,
and they serve as compact sources and detectors for terahertz imaging and
short-range high-data-rate links. The same negative resistance also makes a
fast two-state latch, a memory or logic cell that holds one of two stable
operating points. The device never entered mass logic, because it does not
compose into large circuits as cleanly as the MOSFET, and its story is the
counterweight to the transistor section: the quantum tunnelling that is a
parasitic leak in a scaled gate is, in a deliberately engineered double
barrier, the entire mechanism of a working high-frequency component. The
engineering content is that a quantum effect is neither good nor bad in
itself; whether confinement and tunnelling are a limit to fight or a
mechanism to exploit depends only on which side of the barrier the designer
stands.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-rtd-iv.tex}

\section{An applied vignette: the quantum-cascade laser}\pathtag{mastery}

Every emitter treated so far, the LED and both laser diodes, works by
recombining an electron in the conduction band with a hole in the valence
band, so its photon energy is pinned near the material's band gap and no
ordinary diode reaches far into the infrared. The quantum-cascade laser
breaks that pin, and it is worth a short study because it turns the
particle-in-a-box ladder the chapter has used for wavelength tuning into
the whole gain mechanism, and because a single electron is made to emit
not one photon but tens.

The active region is a stack of tens to hundreds of nominally identical
periods, each a set of coupled quantum wells grown a few nanometres thick
(Figure~\ref{fig:vol04ch12:qcl-staircase}). Confinement splits the
conduction band of each well into subbands, the $E_n = n^{2}\pi^{2}\hbar^{2}
/(2m^{*}L^{2})$ ladder the quantum-well laser used, and the laser
transition is between two of those subbands within the same conduction
band rather than across the gap. The emitted photon energy is the subband
spacing, set by the well widths, so growing thinner wells raises the
spacing and shortens the wavelength while the material is unchanged. A
device built entirely from the GaAs or InP system, whose band gaps sit in
the near infrared, emits anywhere from the mid infrared near
$4\,\si{\micro\meter}$ to the far infrared beyond $100\,\si{\micro\meter}$,
the terahertz band, a reach no interband diode has.

The cascade is the second departure. Under bias the band tilts into a
staircase, and an electron that has dropped between the two subbands of one
period tunnels into the upper subband of the next, drops again, and emits
another photon, repeating once per period down the whole stack. A single
electron therefore yields one photon per period, and a fifty-period device
has an internal quantum efficiency that can exceed unity in photons per
electron, the opposite of the LED where each electron gives at most one
photon. The price is voltage: the applied bias must drop the sum of all the
period voltages, so a mid-infrared quantum-cascade laser runs at ten volts
or more across the stack, and the wall-plug efficiency stays modest because
that voltage carries the electron down the whole staircase whether or not
each drop is radiative.

Two design facts follow from the intersubband nature of the transition.
The upper-subband lifetime is set by fast phonon scattering and is short,
of order a picosecond, far shorter than the nanosecond interband lifetime,
so the device needs a high current density to hold the inversion and runs
hot; early devices worked only pulsed and cooled, and room-temperature
continuous operation in the mid infrared was a materials-and-thermal
achievement rather than a matter of course. The gain spectrum is also
narrow and tunable, which is why the quantum-cascade laser became the
source of choice for trace-gas sensing: tuned to a molecular absorption
line in the mid-infrared fingerprint region, it detects parts per billion
of a target gas in a compact spectrometer. The engineering content is that
confinement, used elsewhere in the chapter only to shift an interband
wavelength, here supplies both the transition and the wavelength, and the
cascade converts the usual one-photon-per-electron ceiling into a
multiplier set by the number of periods.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-qcl-staircase.tex}

\section{An applied vignette: the electro-absorption modulator}\pathtag{mastery}

The laser sections made light and switched it on and off by switching the
current that drives the gain. A high-speed optical link cannot switch the laser
itself that fast, because turning a laser on and off chirps its wavelength as
the carrier density swings the refractive index, so the transmitter runs the
laser continuously and modulates the beam with a separate device. The
electro-absorption modulator is that device, and it is worth a short study
because it turns the quantum-well confinement the chapter used for wavelength
tuning into a voltage-controlled shutter, and because it exposes a quantum
effect the earlier sections had no need of: the field-driven shift of an
absorption edge.

The modulator is a short waveguide whose core is a stack of quantum wells,
placed in the intrinsic region of a reverse-biased pn junction so that a small
applied voltage drops a large field across the wells. The design wavelength sits
just beyond the zero-field absorption edge, in the transparent tail where the
material barely absorbs (Figure~\ref{fig:vol04ch12:eam-absorption}). Applying
reverse bias pulls the electron and hole wavefunctions of each well toward
opposite walls, which lowers the confined-state energies and so shifts the whole
absorption edge to longer wavelength. The operating wavelength that was in the
transparent tail now sits on the rising edge, and the light is absorbed. A swing
of a volt or two therefore switches the transmitted beam between a nearly
transparent on state and a strongly attenuated off state, with no change to the
laser feeding it.

Two figures of merit govern the device. The extinction ratio is the ratio of
the on-state transmission to the off-state transmission through the modulator
length; a stack of ten wells over a $100\,\si{\micro\meter}$ waveguide reaches
ten to fifteen decibels, enough to separate a digital one from a zero. The
insertion loss is the residual absorption in the on state, which the design
minimises by placing the operating wavelength far enough into the tail, in
tension with the extinction ratio that wants it close to the edge. The speed is
where the device wins: the switching is set not by any carrier lifetime but by
the resistance and capacitance of the reverse-biased junction, and a small
capacitance carries the modulator past $40\,\si{\giga\hertz}$ with no chirp
penalty, which is why an electro-absorption-modulated laser, the modulator grown
on the same chip as a continuous-wave distributed-feedback laser, is a standard
transmitter for metro and access optical links. The engineering content is that
the same quantum-well ladder used elsewhere in the chapter to fix an emission
colour, here shifted by a field, becomes a fast low-voltage optical switch, and
the switch is fast because it moves an absorption edge rather than a population
of carriers.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-eam-absorption.tex}

\begin{mastery}
The edge shift the modulator lives on is the quantum-confined Stark effect, and
its field dependence, quadratic rather than linear, is worth carrying because it
explains why the well confines the effect at all. A field $\mathcal{E}$ applied
to an unconfined electron and hole would simply pull them apart until they
ionised, and the transition would smear rather than shift. Confining them to a
well of width $L$ stops the ionisation: the walls hold the carriers together, so
the field only distorts the confined wavefunctions, pushing the electron toward
one wall and the hole toward the other. To first order in the field the energy
shift vanishes because the ground state has no permanent dipole in a symmetric
well; the leading term is second order, the field inducing a dipole $p$
proportional to $\mathcal{E}$ and lowering the energy by
\[
\Delta E \approx -\tfrac{1}{2}\beta\,\mathcal{E}^{2},
\qquad
\beta \sim \frac{e^{2}m^{*}L^{4}}{\hbar^{2}},
\]
a polarisability that grows as the fourth power of the well width. The
fourth-power dependence is the design lever: a wider well shifts far more per
volt, so the extinction ratio per volt rises steeply with well width, while a
wider well also weakens the confinement that keeps the exciton from ionising, so
the well width is optimised, not maximised. The quadratic law also means the
modulator is nonlinear in drive voltage, a chirp-free but distorting transfer
function that the link design must pre-compensate. The engineering content is
that confinement converts a field that would ionise a free exciton into a
clean quadratic energy shift, and the well width sets both how large that shift
is and how much bias the device can take before the exciton is torn apart.
\end{mastery}

\section{An applied vignette: the Josephson junction and the SQUID}\pathtag{mastery}

The tunnelling section sent single electrons through a barrier and the
resonant-tunnelling diode sent them through a well. A Josephson junction
sends paired electrons through a barrier between two superconductors, and
it is worth a short study because it is the chapter's one macroscopic
quantum effect, a current that flows at exactly zero voltage, and because
it underlies both the most sensitive magnetometer built and the definition
of the volt.

Below its critical temperature a superconductor binds its conduction
electrons into Cooper pairs that share a single quantum-mechanical phase
across the whole sample, a condensate described by one wavefunction with a
definite amplitude and phase. Put two superconductors close enough that
their pair wavefunctions overlap through a thin insulating barrier and the
pairs tunnel across as a unit. The current that flows depends only on the
difference $\Delta\phi$ between the two phases,
\[
I = I_c\sin(\Delta\phi),
\]
the DC Josephson relation, and because it flows with no voltage across the
barrier it is a supercurrent, the vertical branch of
Figure~\ref{fig:vol04ch12:josephson-iv}. Drive the current past the
critical value $I_c$ and the junction can no longer carry it as pairs; it
switches to a resistive branch and develops a voltage. Under a voltage $V$
the phase difference winds at a steady rate and the supercurrent
oscillates at the Josephson frequency
\[
f_J = \frac{2eV}{h} = 483.6\,\si{\mega\hertz}\ \text{per}\ \si{\micro\volt},
\]
the factor of two counting the pair charge. That frequency-to-voltage
relation depends only on $e$ and $h$, so a junction irradiated at a known
microwave frequency locks its voltage to an exact multiple of $hf/2e$, and
this is how the volt is now realised in national metrology laboratories, a
fundamental-constant standard rather than a chemical cell.

The device that made the Josephson junction indispensable is the
superconducting quantum interference device, the SQUID. Two junctions in a
superconducting loop form an interferometer for the pair wavefunction: the
magnetic flux threading the loop shifts the relative phase of the two
paths, and because the flux enclosed by a superconducting loop is quantised
in units of the flux quantum $\Phi_0 = h/2e = 2.07\times 10^{-15}\,\si{
\weber}$, the loop's critical current oscillates as the applied flux is
swept, one full oscillation per flux quantum. Reading that oscillation lets
a SQUID resolve a small fraction of a single flux quantum, which for a
practical loop area is a magnetic field below a femtotesla, ten orders of
magnitude finer than a compass and fine enough to map the magnetic fields
of neural currents in the brain (magnetoencephalography) and the heart, and
to serve as the front-end amplifier for axion and dark-matter searches. The
engineering content is that a superconductor's single macroscopic phase
turns quantum mechanics from a small correction at the device scale into
the entire operating principle: the supercurrent, the flux quantum, and the
exact frequency-voltage relation are all direct readings of one shared
wavefunction, and the SQUID is the amplifier that makes that wavefunction's
phase a laboratory instrument.

\begin{mastery}
The flux quantisation the SQUID reads out is a direct consequence of the
single-valuedness of the condensate wavefunction, and the argument is short
enough to carry. The Cooper-pair condensate is described by a wavefunction
$\psi = |\psi|e^{i\theta}$ whose phase $\theta$ must return to itself, up to
a multiple of $2\pi$, on any closed loop, because the wavefunction is
single-valued. Inside a superconductor the current is expelled from the
bulk (the Meissner effect), so deep inside a thick loop the supercurrent is
zero, and where the current vanishes the phase gradient is fixed entirely by
the magnetic vector potential $\mathbf{A}$. Integrating the phase change
around a loop that stays deep in the superconductor,
\[
\oint \nabla\theta\cdot d\mathbf{l}
= \frac{2e}{\hbar}\oint \mathbf{A}\cdot d\mathbf{l}
= \frac{2e}{\hbar}\,\Phi = 2\pi n,
\]
where $\Phi$ is the enclosed flux and the pair charge $2e$ appears because
the condensate is made of pairs. Solving, the flux is quantised,
$\Phi = n\,(h/2e) = n\Phi_0$, in integer units of the flux quantum
$\Phi_0 = h/2e$. The factor of two in the denominator is the experimental
fingerprint of pairing: the measured flux quantum is half what a
single-electron argument would give, which is how the pair charge was
confirmed. A SQUID does not measure flux by counting whole quanta but by
biasing two junctions so their interference converts a fraction of a flux
quantum into a measurable modulation of the critical current, so the
instrument's resolution is set by how small a phase shift the junction pair
can register, not by $\Phi_0$ itself. The engineering content is that a
topological constraint, the single-valued phase of one macroscopic
wavefunction, becomes a metrological ruler finer than any classical field
sensor.
\end{mastery}

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-josephson-iv.tex}

\section{Worked examples}\pathtag{standard}

A run of worked instances fixes the chapter's daily arithmetic: a photon
energy, a diode operating point, a noise budget, a quantum-well
emission wavelength, a tunnel-junction current ratio, a
bandgap-referenced temperature reading, a Hall-effect carrier density,
a MOSFET threshold and drain current, a thermoelectric figure of
merit with the Peltier cooling it supports, a laser mirror loss and
threshold gain, a blackbody peak from Wien's law, an effective
density of states evaluated to a number, an avalanche-photodiode
gain with the excess noise it brings, a WKB transmission evaluated to
an absolute number, a one-sided junction's built-in potential and
depletion width, a quantum-dot emission colour from three-dimensional
confinement, a Stefan-Boltzmann scene-contrast power for a thermal
imager, the Johnson-against-shot-noise crossover that decides which
floor binds a receiver, a varactor tuning ratio, a junction breakdown
voltage split between the Zener and avalanche mechanisms, an LED
junction temperature from a thermal-resistance chain, a detector's
noise-equivalent power and specific detectivity, a photoconductive gain
against the photodiode, a Fowler-Nordheim write field for a flash cell,
the temperature coefficient of a solar cell's open-circuit voltage, a diode
reverse-recovery time from the stored charge, a photodiode junction capacitance
and the bandwidth it sets, an optical-link power budget and its unamplified
reach, an internal quantum efficiency from the ABC recombination rates, a
superconducting energy gap and the temperature it demands, an
electro-absorption modulator's extinction and drive, and an X-ray half-value
layer for imaging and shielding.

\paragraph{Photon energy and the working rule.}
A fibre-optic transmitter runs at $\lambda = 1.55\,\si{\micro\meter}$.
The photon energy is $E = hc/\lambda$. Rather than carry constants, we
use the working rule $E[\si{\electronvolt}] = 1.24 / \lambda[\si{
\micro\meter}] = 1.24/1.55 = 0.80\,\si{\electronvolt}$. The momentum is
$p = h/\lambda = (6.63\times 10^{-34})/(1.55\times 10^{-6}) = 4.28\times
10^{-28}\,\si{\kilogram\meter\per\second}$, and the frequency is
$\nu = c/\lambda = 1.93\times 10^{14}\,\si{\hertz} = 193\,\si{\tera
\hertz}$. The $0.80\,\si{\electronvolt}$ photon energy is below
silicon's $1.12\,\si{\electronvolt}$ gap, which is why silicon is
transparent at telecom wavelengths and why InGaAs, not silicon, is the
photodiode material for fibre links. The arithmetic matches the table
produced by \texttt{code/photon\_energy.py}.

\paragraph{Diode operating point.}
A silicon diode with $I_{s} = 10^{-12}\,\si{\ampere}$ carries a forward
current $I = 1\,\si{\milli\ampere}$. The forward voltage follows from
inverting the Shockley equation: $V = V_{T}\ln(I/I_{s} + 1) \approx
(0.0259)\ln(10^{-3}/10^{-12}) = 0.0259 \times \ln(10^{9}) = 0.0259
\times 20.7 = 0.536\,\si{\volt}$. Raising the current by a factor of
ten to $10\,\si{\milli\ampere}$ adds $V_{T}\ln 10 = 0.0259 \times 2.303
= 60\,\si{\milli\volt}$, the standard decade slope. The forward drop of
a silicon diode is therefore near $0.6$ to $0.7\,\si{\volt}$ across the
ordinary range of operating currents, the working number for every
diode-circuit estimate.

\paragraph{Receiver noise budget.}
A photodiode receiving $P = 1\,\si{\micro\watt}$ at $\lambda = 850\,
\si{\nano\meter}$ with quantum efficiency $\eta = 0.8$ produces a
photocurrent $I = \eta e P \lambda / (hc)$. With $\lambda/(hc) = 1/(1.46
\,\si{\electronvolt}\cdot e) $, the responsivity $\eta e\lambda/(hc) =
0.8 \times 850/1240 = 0.55\,\si{\ampere\per\watt}$, so $I = 0.55\,\si{
\micro\ampere}$. The shot-noise current in a bandwidth $\Delta f = 100
\,\si{\mega\hertz}$ is $\sqrt{2eI\Delta f} = \sqrt{2 \times 1.6\times
10^{-19}\times 5.5\times 10^{-7}\times 10^{8}} = \sqrt{1.76\times
10^{-17}} = 4.2\,\si{\nano\ampere}$. The shot-noise-limited
signal-to-noise ratio is $0.55\,\si{\micro\ampere}/4.2\,\si{\nano
\ampere} \approx 130$, or $42\,\si{\decibel}$, before any Johnson noise
of the load resistor is added. The budget shows the design rule: at
this power the link is comfortably above the shot-noise floor, and the
amplifier's thermal noise, not the photon statistics, will set the
sensitivity.

\paragraph{Quantum-well emission wavelength.}
A GaAs quantum well of width $L = 8\,\si{\nano\meter}$ sits between
AlGaAs barriers. The confinement raises the emission energy above the
bulk GaAs gap $E_{g} = 1.42\,\si{\electronvolt}$ by the ground-state
confinement energy of the electron and the hole, the particle-in-a-box
result $E_{n} = n^{2}\pi^{2}\hbar^{2}/(2m^{*}L^{2})$
(Figure~\ref{fig:vol04ch12:quantum-well-levels}). For the electron with
$m^{*}_{e} = 0.067\,m_{e}$,
\[
E_{1}^{e} = \frac{\pi^{2}(1.055\times 10^{-34})^{2}}{2(0.067 \times
9.11\times 10^{-31})(8\times 10^{-9})^{2}}
= 1.4\times 10^{-20}\,\si{\joule} = 0.088\,\si{\electronvolt}.
\]
The heavy hole with $m^{*}_{h} \approx 0.45\,m_{e}$ has $E_{1}^{h}
\approx 0.013\,\si{\electronvolt}$ by the same formula scaled by the
mass ratio. The emission energy is $E_{g} + E_{1}^{e} + E_{1}^{h} =
1.42 + 0.088 + 0.013 = 1.52\,\si{\electronvolt}$, a wavelength of
$\lambda = 1.24/1.52 = 0.816\,\si{\micro\meter}$, blue-shifted from the
bulk GaAs value of $0.873\,\si{\micro\meter}$. Narrowing the well to
$5\,\si{\nano\meter}$ scales the confinement energies by $(8/5)^{2} =
2.6$, pushing the electron level to $0.23\,\si{\electronvolt}$ and the
emission to about $0.74\,\si{\micro\meter}$. The well width is the
design knob that tunes wavelength at fixed material.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-quantum-well-levels.tex}

\paragraph{Tunnel-junction current ratio.}
A gate oxide is thinned from $d_{1} = 3\,\si{\nano\meter}$ to $d_{2} =
2\,\si{\nano\meter}$ in a process shrink. The tunnelling current
density scales as $J \sim \exp(-2d\sqrt{2m^{*}\phi}/\hbar)$ with barrier
height $\phi = 3.1\,\si{\electronvolt}$ for SiO$_{2}$ and an oxide
electron effective mass $m^{*} \approx 0.5\,m_{e}$. The decay constant
is
\[
\kappa = \frac{2\sqrt{2m^{*}\phi}}{\hbar}
= \frac{2\sqrt{2(0.5\times 9.11\times 10^{-31})(3.1\times 1.6\times
10^{-19})}}{1.055\times 10^{-34}}
= 1.27\times 10^{10}\,\si{\per\meter}.
\]
The current ratio for the $1\,\si{\nano\meter}$ thinning is
\[
\frac{J_{2}}{J_{1}} = \exp[\kappa(d_{1}-d_{2})]
= \exp[(1.27\times 10^{10})(10^{-9})] = \exp(12.7) \approx 3.3\times 10^{5}.
\]
Removing one nanometre of oxide raises the leakage by more than five
orders of magnitude. The exponential sensitivity is why the gate oxide
hit a hard floor near $1.2\,\si{\nano\meter}$ of SiO$_{2}$ and why the
industry moved to physically thicker high-$k$ HfO$_{2}$ that holds the
same capacitance at lower tunnelling, the trend tabulated in
\texttt{code/tunneling\_leakage.py}.

\paragraph{Bandgap-referenced temperature reading.}
A silicon diode driven at a constant forward current of $100\,\si{
\micro\ampere}$ reads a forward voltage $V_{f} = 0.600\,\si{\volt}$ at
$25\,{}^{\circ}\mathrm{C}$. The forward voltage falls by about
$2.0\,\si{\milli\volt}$ per kelvin, the temperature coefficient
dominated by the band-gap temperature dependence inside $I_{s}$ rather
than by the explicit thermal-voltage factor. A later reading of
$V_{f} = 0.560\,\si{\volt}$ corresponds to a temperature change
\[
\Delta T = \frac{V_{f}(25\,{}^{\circ}\mathrm{C}) - V_{f}}{2.0\,\si{
\milli\volt\per\kelvin}}
= \frac{600 - 560\,\si{\milli\volt}}{2.0\,\si{\milli\volt\per\kelvin}}
= 20\,\si{\kelvin},
\]
so the junction sits at $45\,{}^{\circ}\mathrm{C}$. The reading is
linear over a hundred-kelvin span, which is why a two-point calibration
fixes the slope and intercept; the same physics, run in reverse,
underlies the bandgap voltage reference, where the difference of two
junction voltages at a fixed current ratio yields a voltage
proportional to absolute temperature.

\paragraph{Hall-effect carrier density.}
A doped silicon bar of thickness $t = 50\,\si{\micro\meter}$ carries a
current $I = 1\,\si{\milli\ampere}$ in a transverse field $B = 0.5\,\si{
\tesla}$ and develops a Hall voltage $V_{H} = 1.04\,\si{\milli\volt}$
across its width (Figure~\ref{fig:vol04ch12:hall-bar}). The moving
carriers feel the Lorentz force $q\,\mathbf{v}\times\mathbf{B}$ and pile
up on one face until the transverse field they build cancels the
deflection; the steady-state balance gives $V_{H} = I B/(n e t)$.
Solving for the carrier density,
\[
n = \frac{I B}{e\,t\,V_{H}}
= \frac{(10^{-3})(0.5)}{(1.6\times 10^{-19})(5\times 10^{-5})(1.04\times
10^{-3})}
= 6.0\times 10^{22}\,\si{\per\meter\cubed}
= 6.0\times 10^{16}\,\si{\per\centi\meter\cubed}.
\]
The sign of $V_{H}$ fixes the carrier type: a positive Hall voltage on
the face set by the right-hand rule for positive charge marks a p-type
sample, the opposite sign an n-type sample. The number $6\times 10^{16}
\,\si{\per\centi\meter\cubed}$ is a moderate doping, consistent with a
solar-cell base or a lightly doped well. The Hall measurement is the
working route to a device's carrier density and mobility together:
combining $n$ with the bar's resistance gives the mobility $\mu =
1/(n e \rho)$, the two numbers that set every resistive and switching
property of the material.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-hall-bar.tex}

\paragraph{MOSFET threshold and drain current.}
An n-channel MOSFET has gate-oxide thickness $t_{ox} = 2\,\si{\nano
\meter}$, channel mobility $\mu_{n} = 0.04\,\si{\meter\squared\per\volt
\per\second}$, width-to-length ratio $W/L = 10$, threshold voltage
$V_{T} = 0.4\,\si{\volt}$, and runs at $V_{GS} = 1.0\,\si{\volt}$. The
oxide capacitance per unit area sets the gate's hold on the channel
charge,
\[
C_{ox} = \frac{\varepsilon_{0}\varepsilon_{ox}}{t_{ox}}
= \frac{(8.85\times 10^{-12})(3.9)}{2\times 10^{-9}}
= 1.7\times 10^{-2}\,\si{\farad\per\meter\squared}
= 1.7\,\si{\micro\farad\per\centi\meter\squared}.
\]
In saturation ($V_{DS} > V_{GS} - V_{T}$) the drain current is
\[
I_{D} = \tfrac{1}{2}\mu_{n}C_{ox}\frac{W}{L}(V_{GS} - V_{T})^{2}
= \tfrac{1}{2}(0.04)(1.7\times 10^{-2})(10)(0.6)^{2}
= 1.2\times 10^{-3}\,\si{\ampere},
\]
about $1.2\,\si{\milli\ampere}$. The current scales with the square of
the gate overdrive $V_{GS} - V_{T}$, the relation that lets the gate
voltage switch the channel between an off state (a thermally limited
sub-threshold leakage falling about a decade per $60\,\si{\milli\volt}$
of gate swing at room temperature) and an on state carrying milliamperes.
The thinner the oxide, the larger $C_{ox}$ and the harder the gate's
grip on the channel, which is the device pressure that drove $t_{ox}$
down toward the tunnelling floor of the leakage example and forced the
move to high-$k$ dielectrics.

\paragraph{Thermoelectric figure of merit and Peltier cooling.}
A thermoelectric module is built from bismuth-telluride legs with
Seebeck coefficient $S = 200\,\si{\micro\volt\per\kelvin}$, electrical
conductivity $\sigma = 1.0\times 10^{5}\,\si{\siemens\per\meter}$, and
thermal conductivity $k = 1.5\,\si{\watt\per\meter\per\kelvin}$. The
dimensionless figure of merit at $T = 300\,\si{\kelvin}$ is
\[
ZT = \frac{S^{2}\sigma}{k}\,T
= \frac{(2\times 10^{-4})^{2}(1.0\times 10^{5})}{1.5}\,(300)
= \frac{(4\times 10^{-8})(10^{5})}{1.5}(300)
= 0.80.
\]
A figure of merit near $1$ is the working ceiling of commercial
bismuth-telluride modules, and it sets the maximum temperature drop a
single-stage Peltier cooler can hold at zero heat load,
\[
\Delta T_{\max} = \tfrac{1}{2}Z T_{c}^{2}
\approx \tfrac{1}{2}ZT\,T_{c},
\]
which for $ZT = 0.80$ and a cold side near $T_{c} = 270\,\si{\kelvin}$
gives an ideal $\Delta T_{\max} \approx 0.5\times 0.80\times 270 \approx
108\,\si{\kelvin}$; parasitic conduction, contact resistance, and the
finite optimisation of leg geometry and current cut that to the $65$ to
$70\,\si{\kelvin}$ that real single-stage modules reach. The quantum content sits inside the three transport
coefficients: the Seebeck coefficient is the entropy carried per charge,
fixed by the band structure near the Fermi level, and the search for
$ZT > 2$ is a search for materials that scatter phonons (lowering $k$)
without scattering electrons (holding $\sigma$), a decoupling that
nanostructured superlattices and quantum-dot composites pursue. The
Peltier cooler trades the second law against the figure of merit: it
moves heat against a gradient at the cost of $I^{2}R$ dissipation, and
$ZT$ measures how favourably that trade runs.

\paragraph{Laser mirror loss and threshold gain.}
A Fabry-Perot laser cavity is $L = 250\,\si{\micro\meter}$ long with
both facets cleaved bare ($R_1 = R_2 = 0.30$) and an internal loss
$\alpha_i = 12\,\si{\per\centi\meter}$. The mirror loss is the
distributed rate at which power leaves through the facets,
\[
\alpha_m = \frac{1}{2L}\ln\!\left(\frac{1}{R_1 R_2}\right)
         = \frac{1}{2(0.025\,\si{\centi\meter})}\ln\!\left(
           \frac{1}{0.09}\right)
         = 20\times 2.41 = 48\,\si{\per\centi\meter},
\]
so the threshold gain is $g_{th} = \alpha_i + \alpha_m = 60\,\si{\per
\centi\meter}$. Coating the rear facet to $R_2 = 0.95$ changes the
logarithm to $\ln(1/(0.30\times 0.95)) = \ln(3.51) = 1.26$, dropping
$\alpha_m$ to $20\times 1.26 = 25\,\si{\per\centi\meter}$ and the
threshold gain to $37\,\si{\per\centi\meter}$, a $38\,\%$ cut. The
useful-output fraction $\alpha_m/(\alpha_i+\alpha_m)$ falls from $0.80$
to $0.68$, so the rear coating trades a lower threshold against a slightly
lower slope efficiency, the design choice that sets how a given die is
binned for low-power or high-power use.

\paragraph{Blackbody peak from Wien's law.}
A tungsten lamp filament runs at $T = 2800\,\si{\kelvin}$. Wien's
displacement law places the spectral-radiance peak at
\[
\lambda_{\max} = \frac{b}{T}
              = \frac{2.898\times 10^{-3}\,\si{\meter\kelvin}}
                     {2800\,\si{\kelvin}}
              = 1.04\times 10^{-6}\,\si{\meter}
              = 1.04\,\si{\micro\meter},
\]
in the near infrared, which is why an incandescent lamp spends most of
its radiated power as heat rather than visible light and why its
luminous efficacy is only about $15\,\si{\lumen\per\watt}$ against the
white-LED luminaire's $110$. Raising the filament to the $3680\,\si{
\kelvin}$ just below tungsten's melting point moves the peak to
$0.79\,\si{\micro\meter}$, still in the infrared; no attainable filament
temperature pushes the peak into the green where the eye is most
sensitive, which is the thermodynamic reason solid-state emission
replaced the hot filament. The peak shifts inversely with temperature,
the displacement seen in Figure~\ref{fig:vol04ch12:blackbody-curves}.

\paragraph{Effective density of states evaluated to a number.}
The effective density of states $N_c = 2(m^{*}_e k_B T/2\pi\hbar^{2})^{
3/2}$ derived above is worth carrying to a number for silicon at
$300\,\si{\kelvin}$ with conduction-band effective mass $m^{*}_e =
0.26\,m_e$ (the simple-parabolic value; the multi-valley value differs).
The bracket is
\[
\frac{m^{*}_e k_B T}{2\pi\hbar^{2}}
 = \frac{(0.26\times 9.11\times 10^{-31})(1.38\times 10^{-23})(300)}
        {2\pi(1.055\times 10^{-34})^{2}}
 = 1.40\times 10^{16}\,\si{\per\meter\squared},
\]
and raising to the $3/2$ power and multiplying by two gives $N_c =
2(1.40\times 10^{16})^{3/2} = 2\times 1.66\times 10^{24} = 3.3\times
10^{24}\,\si{\per\meter\cubed} = 3.3\times 10^{18}\,\si{\per\centi
\meter\cubed}$, within a factor of the tabulated $2.8\times 10^{19}\,\si{
\per\centi\meter\cubed}$ once the multi-valley degeneracy and the
density-of-states effective mass are folded in. The point of the estimate
is the scale: $N_c$ is of order $10^{18}$ to $10^{19}\,\si{\per\centi
\meter\cubed}$, so the intrinsic density $n_i = \sqrt{N_c N_v}\exp(-E_g/
2k_B T) \approx 10^{20}\exp(-21.6) \approx 10^{10}\,\si{\per\centi\meter
\cubed}$ recovers the value used throughout the chapter, the exponential
of the half-gap doing all the work.

\paragraph{Avalanche-photodiode gain and excess noise.}
An avalanche photodiode (APD) receives a primary photocurrent $I_p =
0.2\,\si{\micro\ampere}$ and is biased to an avalanche gain $M = 50$. The
output current is $M I_p = 10\,\si{\micro\ampere}$, lifting the signal far
above a downstream amplifier's noise. The avalanche multiplication is
itself a random process, so the shot noise is amplified by more than the
gain: the noise spectral density is $2eI_p M^{2}F$, where $F$ is the
excess-noise factor. For a device with ionisation ratio $k = 0.3$,
\[
F = k M + (1-k)\left(2 - \frac{1}{M}\right)
  = 0.3\times 50 + 0.7\times(2 - 0.02)
  = 15 + 1.39 = 16.4.
\]
The output shot-noise current in a bandwidth $\Delta f = 100\,\si{\mega
\hertz}$ is $\sqrt{2eI_p M^{2}F\,\Delta f} = M\sqrt{2eI_p F\,\Delta f} =
50\sqrt{2(1.6\times 10^{-19})(2\times 10^{-7})(16.4)(10^{8})} =
50\times 1.0\times 10^{-8} = 0.52\,\si{\micro\ampere}$. The signal-to-noise
ratio is $10/0.52 \approx 19$, against a unity-gain photodiode whose
amplifier noise would have buried the $0.2\,\si{\micro\ampere}$ primary
signal entirely. The lesson is that avalanche gain helps only until the
multiplied shot noise overtakes the amplifier noise; past the optimum
gain the excess-noise factor $F$, growing roughly as $M$ for this $k$,
makes more gain a net loss, the trade behind the judgment exercise on
photodiode choice.

\paragraph{WKB transmission evaluated to a number.}
A scanning-tunnelling-microscope tip sits in vacuum a distance $d =
0.5\,\si{\nano\meter}$ from a metal surface; the barrier the electron
sees is the work function, about $\phi = 4.5\,\si{\electronvolt}$, and
the mass is the free-electron mass. The decay constant from the WKB
exponent derived above is
\[
\kappa = \frac{\sqrt{2m_e\phi}}{\hbar}
= \frac{\sqrt{2(9.11\times 10^{-31})(4.5\times 1.6\times 10^{-19})}}
       {1.055\times 10^{-34}}
= 1.09\times 10^{10}\,\si{\per\meter},
\]
so the transmission is $T \approx \exp(-2\kappa d) = \exp(-2\times
1.09\times 10^{10}\times 0.5\times 10^{-9}) = \exp(-10.9) = 1.8\times
10^{-5}$. Pulling the tip back by one further \aa ngstr\"om to $d =
0.6\,\si{\nano\meter}$ multiplies the exponent by $1.2$ to $-13.1$, so
$T = 2.0\times 10^{-6}$, a factor of nine for one \aa ngstr\"om. The
exponential sensitivity is the whole working principle of the
scanning-tunnelling microscope: the tunnelling current tracks the gap
so steeply that holding it constant maps the surface to atomic height
resolution, and the same $\kappa$ near $10^{10}\,\si{\per\meter}$ for a
few-electronvolt barrier is the number that recurs across the gate-oxide
and flash-memory examples.

\paragraph{One-sided junction: built-in potential and depletion width.}
A silicon photodiode is built as a one-sided abrupt junction, a heavily
doped n$^{+}$ region ($N_D = 10^{19}\,\si{\per\centi\meter\cubed}$) on a
lightly doped p base ($N_A = 10^{15}\,\si{\per\centi\meter\cubed}$). The
built-in potential is
\[
V_{bi} = \frac{k_B T}{e}\ln\!\left(\frac{N_A N_D}{n_i^{2}}\right)
= 0.0259\,\ln\!\left(\frac{10^{15}\times 10^{19}}{(10^{10})^{2}}\right)
= 0.0259\,\ln(10^{14}) = 0.0259\times 32.2 = 0.83\,\si{\volt}.
\]
Because $N_D \gg N_A$, the depletion region sits almost entirely in the
lightly doped p base, $x_n \approx 0$ and $W \approx x_p$, and the width
formula collapses to the one-sided result
\[
W \approx \sqrt{\frac{2\varepsilon V_{bi}}{eN_A}}
= \sqrt{\frac{2(11.7\times 8.85\times 10^{-12})(0.83)}
             {(1.6\times 10^{-19})(10^{21})}}
= 1.0\times 10^{-6}\,\si{\meter} = 1.0\,\si{\micro\meter},
\]
using $N_A = 10^{15}\,\si{\per\centi\meter\cubed} = 10^{21}\,\si{\per
\meter\cubed}$ and silicon's relative permittivity $11.7$. A reverse bias
of $10\,\si{\volt}$ widens this to $W = \sqrt{2\varepsilon(V_{bi}+10)/
eN_A} \approx 3.6\,\si{\micro\meter}$, since the width grows as the square
root of total junction voltage. The wide, lightly doped depletion region
is exactly what a photodiode wants: a thick collection volume in which
above-gap photons generate carriers inside the field, and the reverse bias
both widens it and speeds the carrier transit, the design choice behind
the PIN photodiode of the receiver examples.

\paragraph{Quantum-dot emission colour from three-dimensional confinement.}
A spherical CdSe quantum dot of diameter $L = 4\,\si{\nano\meter}$ is
modelled as an electron and a hole each confined in a three-dimensional
box of side $L$. The confinement raises the emission energy above the bulk
CdSe gap $E_g = 1.74\,\si{\electronvolt}$ by the ground-state energies of
both carriers, the three-dimensional particle-in-a-box result $E_{111} =
3\pi^{2}\hbar^{2}/(2m^{*}L^{2})$, the factor of three counting the three
confined directions. With a reduced effective mass for the electron-hole
pair near $m^{*} \approx 0.1\,m_e$,
\[
E_{\text{conf}} = \frac{3\pi^{2}(1.055\times 10^{-34})^{2}}
                       {2(0.1\times 9.11\times 10^{-31})(4\times 10^{-9})^{2}}
= 1.13\times 10^{-19}\,\si{\joule} = 0.71\,\si{\electronvolt}.
\]
The emission energy is $E_g + E_{\text{conf}} = 1.74 + 0.71 = 2.45\,\si{
\electronvolt}$, a wavelength $\lambda = 1.24/2.45 = 0.51\,\si{\micro
\meter}$, green. Shrinking the dot to $3\,\si{\nano\meter}$ scales the
confinement by $(4/3)^{2} = 1.8$ to $1.26\,\si{\electronvolt}$ and the
emission to $1.74 + 1.26 = 3.0\,\si{\electronvolt}$, $\lambda = 0.41\,\si{
\micro\meter}$, blue; growing it to $6\,\si{\nano\meter}$ drops the
confinement to $0.32\,\si{\electronvolt}$ and pushes the emission toward
red. The dot diameter is the single knob that tunes colour across the
visible at fixed material, the property exploited in quantum-dot displays
and the three-dimensional analogue of the quantum-well width knob.

\paragraph{Stefan-Boltzmann scene contrast for a thermal imager.}
The microbolometer vignette claimed a scene-contrast power that a
millikelvin membrane rise must register. The radiated power per unit area
from a blackbody scene at $T = 300\,\si{\kelvin}$ changes with scene
temperature at the rate
\[
\frac{dM}{dT} = 4\sigma T^{3}
= 4(5.67\times 10^{-8})(300)^{3}
= 6.1\,\si{\watt\per\meter\squared\per\kelvin},
\]
so a one-kelvin warmer target radiates an extra $6.1\,\si{\watt\per
\meter\squared}$ into the hemisphere. An $f/1$ lens collects a fraction of
that of order $1/(4 f^{2}) = 0.25$, and a $17\,\si{\micro\meter}$ pixel
subtends an area $A = (17\times 10^{-6})^{2} = 2.9\times 10^{-10}\,\si{
\meter\squared}$, so the change in power on one pixel for a one-kelvin
scene change is about $P \approx 6.1\times 0.25\times 0.7\times 2.9\times
10^{-10} \approx 3\times 10^{-10}\,\si{\watt}$, taking a $0.7$ optical
transmission. With the pixel thermal conductance $G = 10^{-7}\,\si{\watt
\per\kelvin}$ the membrane rise is $\Delta T = P/G \approx 3\,\si{\milli
\kelvin}$ per kelvin of scene contrast, the millikelvin figure the
vignette used, and the reason the readout must resolve resistance changes
of parts in $10^{5}$. The estimate also shows the design lever: a lower
$G$, from longer or thinner legs, raises the membrane rise per watt and
the sensitivity, at the cost of a slower thermal time constant.

\paragraph{Johnson-against-shot-noise crossover in a receiver.}
A photodiode delivers a signal photocurrent $I$ into a load resistor $R$
that feeds the amplifier; the shot noise of the photocurrent and the
Johnson noise of the load compete, and the crossover decides which floor
binds. The shot-noise current variance is $2eI\Delta f$ and the Johnson
current variance of the load is $4k_B T\Delta f/R$. Setting them equal,
the photocurrent at which the two are comparable is
\[
I_{\text{cross}} = \frac{2k_B T}{eR} = \frac{2V_T}{R},
\qquad
2V_T = 2\times 26\,\si{\milli\volt} = 52\,\si{\milli\volt},
\]
with the thermal voltage $V_T = k_B T/e = 26\,\si{\milli\volt}$, so the
crossover current is $52\,\si{\milli\volt}$ divided by the load. For a
$1\,\si{\kilo\ohm}$ load this is $52\,\si{\micro\ampere}$: below this
photocurrent Johnson noise dominates and a larger load resistor helps;
above it shot noise dominates and a larger resistor no longer buys
sensitivity. For a $1\,\si{\mega\ohm}$ transimpedance the crossover drops
to $52\,\si{\nano\ampere}$, which is why high-sensitivity receivers run
the largest load the bandwidth allows: raising $R$ pushes the Johnson
floor down until the unavoidable shot noise of the signal itself becomes
the limit, the boundary the failure section names as the quantum floor.

\paragraph{Varactor capacitance and the tuning ratio.}
The depletion width narrows and widens with bias, so the junction is a
voltage-controlled capacitor. For an abrupt junction the depletion charge
per unit area is $Q = eN_D x_n$ and the small-signal capacitance is $C_j =
\varepsilon/W$, with the width growing as the square root of the total
junction voltage $V_{bi}+V_R$. The capacitance therefore falls as
\[
C_j(V_R) = \frac{C_{j0}}{\sqrt{1 + V_R/V_{bi}}},
\qquad
C_{j0} = \frac{\varepsilon}{W_0},
\]
the abrupt-junction law plotted in
Figure~\ref{fig:vol04ch12:varactor-cv}. Take a diode with zero-bias
capacitance $C_{j0} = 10\,\si{\pico\farad}$ and $V_{bi} = 0.83\,\si{\volt}$.
At $V_R = 4\,\si{\volt}$ the capacitance is $10/\sqrt{1 + 4/0.83} =
10/\sqrt{5.82} = 4.1\,\si{\pico\farad}$; at $V_R = 16\,\si{\volt}$ it is
$10/\sqrt{1 + 16/0.83} = 10/\sqrt{20.3} = 2.2\,\si{\pico\farad}$, a tuning
ratio of $1.9$ across the $12\,\si{\volt}$ swing. Because a resonant tank
tunes as $f \propto 1/\sqrt{C}$, that capacitance ratio moves the oscillator
frequency by $\sqrt{1.9} = 1.4$, a $40\,\%$ tuning band. A hyperabrupt
doping profile, graded so the lightly doped side thins toward the junction,
steepens the $C$-$V$ law toward $C \propto (V_{bi}+V_R)^{-1}$ and widens the
tuning ratio, the design used where a single control voltage must cover a
broad band. The varactor is the pn junction's depletion electrostatics sold
as a component, the same Poisson-equation result the junction section carried
once.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-varactor-cv.tex}

\paragraph{Junction breakdown: Zener against avalanche.}
Reverse bias widens the depletion region and raises the peak field, and past
a critical field the reverse current runs away in breakdown. The peak field
of a one-sided junction into a base of doping $N$ is $E_{\max} =
\sqrt{2eN(V_{bi}+V_R)/\varepsilon}$, so the breakdown voltage is the reverse
bias that lifts $E_{\max}$ to the material's critical field $E_{crit}$,
\[
V_{BR} \approx \frac{\varepsilon E_{crit}^{2}}{2eN}.
\]
For silicon with $E_{crit} \approx 3\times 10^{5}\,\si{\volt\per\centi
\meter} = 3\times 10^{7}\,\si{\volt\per\meter}$ and a base doping $N =
10^{16}\,\si{\per\centi\meter\cubed} = 10^{22}\,\si{\per\meter\cubed}$,
\[
V_{BR} = \frac{(11.7\times 8.85\times 10^{-12})(3\times 10^{7})^{2}}
              {2(1.6\times 10^{-19})(10^{22})}
       = \frac{9.3\times 10^{4}}{3.2\times 10^{3}}
       \approx 29\,\si{\volt}.
\]
Because $V_{BR}$ scales inversely with doping, a lightly doped junction
stands off hundreds of volts while a heavily doped one breaks down below ten,
the trade behind every diode's voltage rating. The mechanism sorts by
voltage. Below about $5\,\si{\volt}$ the junction is so heavily doped and the
depletion region so thin that the field tears electrons straight across the
gap by band-to-band tunnelling, the Zener mechanism, whose breakdown voltage
falls with temperature. Above about $6\,\si{\volt}$ the carriers gain enough
energy between collisions to ionise the lattice by impact, the avalanche
mechanism, whose breakdown voltage rises with temperature because hotter
lattice vibrations shorten the mean free path. The opposite temperature signs
are how a bench measurement tells the two apart, and a diode designed near
$5.6\,\si{\volt}$, where the two cancel, is the temperature-stable reference
that preceded the bandgap reference.

\paragraph{LED junction temperature from a thermal balance.}
The white-LED case study drove a die at nearly $10\,\si{\watt}$ with only a
quarter leaving as light, so the heat-sink question is a thermal balance
worth carrying to a number. Model the path from junction to ambient as a
chain of thermal resistances in series, junction-to-case $R_{jc} = 4\,\si{
\kelvin\per\watt}$, case-to-heat-sink $R_{cs} = 1\,\si{\kelvin\per\watt}$,
and heat-sink-to-ambient $R_{sa} = 3\,\si{\kelvin\per\watt}$, a total
$R_{ja} = 8\,\si{\kelvin\per\watt}$. The die dissipates the input power less
the light that leaves, here $P_{th} = 9.6 - 2.4 = 7.2\,\si{\watt}$ of heat.
The junction temperature above a $35\,{}^{\circ}\mathrm{C}$ ambient is
\[
T_j = T_a + P_{th}R_{ja} = 35 + (7.2)(8) = 35 + 58 = 93\,{}^{\circ}
\mathrm{C},
\]
above the $85\,{}^{\circ}\mathrm{C}$ target the case study assumed. The fix
is read straight off the chain: dropping the heat-sink resistance $R_{sa}$
from $3$ to $1.5\,\si{\kelvin\per\watt}$ with a larger fin area cuts $R_{ja}$
to $6.5\,\si{\kelvin\per\watt}$ and $T_j$ to $35 + 7.2\times 6.5 =
82\,{}^{\circ}\mathrm{C}$, inside target. The balance shows why the heat sink
is the binding constraint: junction temperature sets both the efficacy
derating and, through the Arrhenius acceleration of the fade mechanisms, the
L70 lifetime, so every kelvin bought at the fin is bought twice.

\paragraph{Noise-equivalent power and specific detectivity of a detector.}
A detector's sensitivity is quoted as the optical power that produces a
signal equal to the noise in unit bandwidth, the \engterm{noise-equivalent
power} (NEP), and as the area-normalised inverse of it, the \engterm{specific
detectivity} $D^{*} = \sqrt{A\,\Delta f}/\text{NEP}$
(Figure~\ref{fig:vol04ch12:detectivity-spectrum}). Take a silicon photodiode
of area $A = 1\,\si{\milli\meter\squared} = 10^{-2}\,\si{\centi\meter
\squared}$, responsivity $\mathcal{R} = 0.6\,\si{\ampere\per\watt}$ at
$850\,\si{\nano\meter}$, and a dark current $I_d = 1\,\si{\nano\ampere}$
whose shot noise sets the floor. The dark-shot noise-current density is
$i_n = \sqrt{2eI_d} = \sqrt{2(1.6\times 10^{-19})(10^{-9})} = 1.8\times
10^{-14}\,\text{A}/\sqrt{\text{Hz}}$. The NEP is that noise current
divided by the responsivity,
\[
\text{NEP} = \frac{i_n}{\mathcal{R}}
           = \frac{1.8\times 10^{-14}}{0.6}
           = 3.0\times 10^{-14}\,\text{W}/\sqrt{\text{Hz}},
\]
so detecting $30\,\si{\femto\watt}$ in a one-hertz bandwidth gives unit
signal-to-noise. The specific detectivity is
\[
D^{*} = \frac{\sqrt{A}}{\text{NEP}}
      = \frac{\sqrt{10^{-2}}}{3.0\times 10^{-14}}
      = \frac{0.1}{3.0\times 10^{-14}}
      = 3.3\times 10^{12}\,\text{cm}\,\sqrt{\text{Hz}}/\text{W},
\]
in the $10^{12}$ band that marks a good room-temperature photodiode. The
point of $D^{*}$ is that it strips out the detector area, so a small clean
diode and a large one of the same material land on the same curve and the
choice of detector reduces to reading which curve peaks at the working
wavelength.

\begin{mastery}
The detectivity ceiling is not set by the detector but by the light it is
asked to see, and the transition is worth stating because it decides when
cooling stops helping. Every detector carries a dark-noise floor, the shot
noise of its own leakage current, which cooling lowers because the leakage
$\sim n_i \sim \exp(-E_g/2k_B T)$ falls steeply with temperature; this is why
long-wave infrared detectors of narrow gap must be cooled to reach their
rated $D^{*}$. As the dark noise is cooled away, a second floor appears that
no cooling removes: the shot noise of the background photon flux the detector
sees, the thermal radiation of the scene and the instrument itself. A
detector whose noise is set by this background flux is
\engterm{background-limited} (the BLIP condition), and its detectivity is
\[
D^{*}_{\text{BLIP}} = \frac{\lambda}{hc}
\sqrt{\frac{\eta}{2\,\Phi_b}},
\]
with $\eta$ the quantum efficiency and $\Phi_b$ the background photon flux per
unit area. Past this point the only gains are a colder or narrower field of
view that lets fewer background photons in, or a smaller optical bandwidth,
not a better detector. The engineering content is that a detector
specification has two regimes, a dark-limited one where cooling and cleaner
material pay, and a background-limited one where the scene sets the floor and
the design moves to the optics, and the same shot-noise argument that governs
the photodiode receiver draws the line between them.
\end{mastery}

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-detectivity-spectrum.tex}

\paragraph{Photoconductive gain against the photodiode.}
A photodiode delivers at most one electron per absorbed photon, a gain of
one; a photoconductor can deliver many, and the reason is a lifetime-transit
ratio worth carrying. In a photoconductor an absorbed photon frees a
carrier that drifts to a contact under the applied field and is replaced from
the other contact to keep the sample neutral, and it keeps circulating until
it recombines. The gain is the ratio of the carrier lifetime $\tau$ to the
transit time $t_{tr} = L^{2}/(\mu V)$ across a bar of length $L$ under
voltage $V$,
\[
G = \frac{\tau}{t_{tr}} = \frac{\tau\mu V}{L^{2}}.
\]
For a film with carrier lifetime $\tau = 1\,\si{\milli\second}$, mobility
$\mu = 0.01\,\si{\meter\squared\per\volt\per\second}$, electrode spacing
$L = 10\,\si{\micro\meter} = 10^{-5}\,\si{\meter}$, and $V = 5\,\si{\volt}$,
\[
G = \frac{(10^{-3})(0.01)(5)}{(10^{-5})^{2}}
  = \frac{5\times 10^{-5}}{10^{-10}}
  = 5\times 10^{5},
\]
half a million electrons per photon, which is why photoconductors are chosen
where raw responsivity matters more than speed or noise. The price is in both:
the gain-bandwidth product is capped because a long lifetime that raises the
gain also slows the response to $\sim\tau$, and the same generation-
recombination that provides the gain adds its own noise, so the
photoconductor trades the photodiode's unity-gain quantum cleanliness for
current gain. The choice between the two is the choice between a fast, quiet,
low-gain junction and a slow, noisy, high-gain resistor, the detector version
of the avalanche-gain trade.

\paragraph{Fowler-Nordheim field and the flash write path.}
The flash-memory example claimed that a field tilts the oxide barrier and
turns the ten-year retention barrier into a microsecond write path; the
Fowler-Nordheim law makes the claim a number. Under a field $\mathcal{E}$
the rectangular oxide barrier of height $\phi$ becomes a triangle, and the
electron need only tunnel through the width $\phi/(e\mathcal{E})$ at which the
tilted barrier drops to the electron energy. Substituting a linear $V(x) =
\phi - e\mathcal{E}x$ into the WKB integral of the transmission section and
evaluating gives the Fowler-Nordheim tunnelling distance and current
\[
J_{FN} = A\,\mathcal{E}^{2}\exp\!\left(-\frac{B}{\mathcal{E}}\right),
\qquad
B = \frac{4\sqrt{2m^{*}}\,\phi^{3/2}}{3e\hbar},
\]
the field in the exponent rather than the thickness. For SiO$_2$ with
$\phi = 3.1\,\si{\electronvolt}$ and $m^{*} = 0.5\,m_e$ the field constant
is $B \approx 2.4\times 10^{10}\,\si{\volt\per\meter}$. Program an $8\,\si{
\nano\meter}$ oxide by dropping $12\,\si{\volt}$ across it, a field
$\mathcal{E} = 12/(8\times 10^{-9}) = 1.5\times 10^{9}\,\si{\volt\per
\meter}$, and the exponent is $-B/\mathcal{E} = -2.4\times 10^{10}/1.5\times
10^{9} = -16$; drop the field to the $\sim 3\times 10^{7}\,\si{\volt\per
\meter}$ of ordinary retention and the exponent becomes $-800$, an
astronomically smaller current. The $784$-unit swing in the exponent between
write and retention is the ten-years-against-microseconds ratio the flash
cell lives on, and it is the same WKB exponent read as a function of field
rather than thickness.

\paragraph{Temperature coefficient of the open-circuit voltage.}
A solar array on a hot roof delivers less voltage than its datasheet, and the
loss is governed by the temperature dependence of $V_{oc}$, worth carrying
because it dominates the array's derating. The open-circuit voltage is
$V_{oc} = (k_B T/e)\ln(J_{sc}/J_0)$, and although the prefactor $k_B T/e$
rises with temperature, the saturation current $J_0 \propto n_i^{2} \propto
T^{3}\exp(-E_g/k_B T)$ climbs far faster, so $V_{oc}$ falls. Differentiating,
the temperature coefficient is
\[
\frac{dV_{oc}}{dT} = -\frac{E_g/e - V_{oc} + (3k_B T/e)}{T}.
\]
For silicon at $T = 300\,\si{\kelvin}$ with $E_g/e = 1.12\,\si{\volt}$,
$V_{oc} = 0.70\,\si{\volt}$, and $3k_B T/e = 0.078\,\si{\volt}$,
\[
\frac{dV_{oc}}{dT}
= -\frac{1.12 - 0.70 + 0.078}{300}
= -\frac{0.498}{300}
= -1.7\,\si{\milli\volt\per\kelvin},
\]
close to the datasheet $-2\,\si{\milli\volt}$ per kelvin per cell. A module
of $60$ cells in series therefore loses about $60\times 2 = 120\,\si{\milli
\volt\per\kelvin}$, so a panel that runs $30\,\si{\kelvin}$ above its rated
$25\,{}^{\circ}\mathrm{C}$ on a hot day gives up $3.6\,\si{\volt}$ and a
corresponding several percent of power. The formula also shows the material
lever: a wider-gap cell has a smaller $E_g/e - V_{oc}$ head-room and so a
gentler temperature coefficient, one more reason III-V cells hold up better
in the heat than silicon.

\paragraph{Photon flux from a laser pointer.}
A green laser pointer emits $P = 1\,\si{\milli\watt}$ at $\lambda = 532\,\si{
\nano\meter}$. The photon energy is $E = 1.24/0.532 = 2.33\,\si{
\electronvolt} = 2.33\times 1.6\times 10^{-19} = 3.73\times 10^{-19}\,\si{
\joule}$, so the photon emission rate is
\[
\dot N = \frac{P}{E}
       = \frac{10^{-3}}{3.73\times 10^{-19}}
       = 2.7\times 10^{15}\,\si{\per\second}.
\]
Nearly three thousand million million photons a second leave the aperture,
the count so large that the beam is completely classical: its shot noise,
the square root of the count, is $\sqrt{2.7\times 10^{15}} = 5\times
10^{7}\,\si{\per\second}$, a fractional fluctuation of $2\times 10^{-8}$
that no eye or ordinary detector registers. Attenuate the same beam by
$15$ orders of magnitude, to $10^{-18}\,\si{\watt}$, and the rate falls to
about three photons a second, where the arrivals become countable and the
Poisson statistics of the failure section dominate. The example fixes the
scale that separates the classical-light regime of the pointer from the
photon-counting regime of the single-photon detector: both are the same
photons, and only the flux decides which physics binds.

\paragraph{Responsivity ceiling and the quantum-efficiency shortfall.}
A photodiode's responsivity is the photocurrent per watt of light, and its
ceiling is the one-electron-per-photon limit. At unit quantum efficiency the
responsivity is $\mathcal{R} = e\lambda/(hc) = \lambda[\si{\micro\meter}]/
1.24\,\si{\ampere\per\watt}$, rising linearly with wavelength because a
longer-wavelength photon carries less energy so a fixed power delivers more
photons. At $\lambda = 900\,\si{\nano\meter}$ the ceiling is $0.900/1.24 =
0.73\,\si{\ampere\per\watt}$. A real silicon device there measures
$\mathcal{R} = 0.55\,\si{\ampere\per\watt}$, so its external quantum
efficiency is $\eta = 0.55/0.73 = 0.75$: three photons in four make a
collected carrier, the shortfall going to front-surface reflection, to
photons passing through the thin absorber near the band edge, and to
carriers that recombine before collection. The responsivity ceiling falls
abruptly to zero past the band-edge wavelength $\lambda_g = 1.24/E_g =
1.11\,\si{\micro\meter}$ for silicon, which is why a silicon photodiode
peaks near $900\,\si{\nano\meter}$ and is blind to the $1.55\,\si{\micro
\meter}$ telecom band, the reason the earlier photon-energy example put
InGaAs, not silicon, on the fibre link.

\paragraph{Debye length and the screening of a junction.}
The depletion approximation drew the space charge as a sharp step, but the
edge is smeared over a screening length worth carrying because it sets when
the abrupt-junction picture holds. Mobile carriers rearrange to screen a
disturbance over the \engterm{Debye length}
\[
L_D = \sqrt{\frac{\varepsilon k_BT}{e^{2}N}},
\]
the distance over which a field falls by $1/e$ in a doped semiconductor. For
silicon ($\varepsilon = 11.7\times 8.85\times 10^{-12}\,\si{\farad\per
\meter}$) doped to $N = 10^{16}\,\si{\per\centi\meter\cubed} = 10^{22}\,\si{
\per\meter\cubed}$,
\[
L_D = \sqrt{\frac{(1.04\times 10^{-10})(1.38\times 10^{-23})(300)}
                 {(1.6\times 10^{-19})^{2}(10^{22})}}
    = \sqrt{1.68\times 10^{-15}}
    = 4.1\times 10^{-8}\,\si{\meter}
    = 41\,\si{\nano\meter}.
\]
A depletion width of $150\,\si{\nano\meter}$ at this doping is only a few
Debye lengths wide, so the sharp-step depletion approximation is a fair but
not exact idealisation: the true charge edge is graded over the Debye
length, and the correction matters most for lightly doped, low-voltage
junctions where the width and the screening length become comparable. The
same length screens a contact potential, sets the thickness of an
accumulation layer, and fixes how closely two junctions may be placed before
their space-charge regions merge, the short-channel limit of the transistor.

\paragraph{Single-photon detector and the dead-time correction.}
A single-photon avalanche diode counts individual photons but is blinded
for a dead time $\tau = 40\,\si{\nano\second}$ after each count, so its
measured rate saturates (Figure~\ref{fig:vol04ch12:deadtime-saturation}).
For the non-paralysable model the measured rate $m$ relates to the true rate
$r$ by $m = r/(1 + r\tau)$, inverted to recover the truth,
\[
r = \frac{m}{1 - m\tau}.
\]
A detector reporting $m = 5\times 10^{6}\,\si{\per\second}$ has $m\tau =
(5\times 10^{6})(40\times 10^{-9}) = 0.20$, so the true rate is $r =
5\times 10^{6}/(1 - 0.20) = 6.25\times 10^{6}\,\si{\per\second}$, a $25\,\%$
undercount that the correction restores. The maximum measured rate the
detector can report is its saturation ceiling $1/\tau = 2.5\times 10^{7}\,
\si{\per\second}$, approached only at infinite true flux, and the count
becomes unreliable once $m\tau$ nears unity because the correction divides
by a small number. The working rule is to keep $m\tau$ below about $0.1$, a
true rate under $2.8\times 10^{6}\,\si{\per\second}$ for this dead time, so
the correction stays a small perturbation rather than a large extrapolation.

\paragraph{Josephson voltage and the frequency standard.}
A Josephson junction irradiated at a microwave frequency $f = 70\,\si{\giga
\hertz}$ develops quantised voltage steps at integer multiples of $hf/2e$.
The first step sits at
\[
V_1 = \frac{hf}{2e}
    = \frac{(6.63\times 10^{-34})(70\times 10^{9})}{2(1.6\times 10^{-19})}
    = 1.45\times 10^{-4}\,\si{\volt}
    = 145\,\si{\micro\volt},
\]
using the inverse Josephson relation $f_J = 2eV/h$. Because the step voltage
depends only on the applied frequency and the fundamental constants $e$ and
$h$, a chip carrying many thousands of junctions in series, each on the
same step, sums to a volt-level standard reproducible anywhere to parts in
$10^{10}$, far beyond any chemical cell. To realise exactly one volt a
series array must supply $1/(145\times 10^{-6}) \approx 6900$ junctions on
the first step at this drive frequency. The example is the metrology payoff
of the Josephson vignette: the frequency-to-voltage conversion is exact, so
frequency, the quantity engineers measure best, defines the volt.

\paragraph{Peltier cooler coefficient of performance.}
The thermoelectric worked example found the maximum temperature drop of a
Peltier module; the useful figure for a cooler holding a load is its
coefficient of performance, the heat pumped per unit electrical power. For a
module of figure of merit $ZT$ pumping across a small temperature difference
$\Delta T$ between a cold side $T_c$ and hot side $T_h$, the optimum
coefficient of performance is
\[
\mathrm{COP} = \frac{T_c}{\Delta T}\cdot
\frac{\sqrt{1 + Z\bar T} - T_h/T_c}{\sqrt{1 + Z\bar T} + 1},
\]
with $\bar T$ the mean temperature. Take $ZT = 0.8$ so $\sqrt{1 + Z\bar T} =
\sqrt{1.8} = 1.34$, a cold side $T_c = 283\,\si{\kelvin}$ and hot side
$T_h = 303\,\si{\kelvin}$ ($\Delta T = 20\,\si{\kelvin}$), giving
\[
\mathrm{COP} = \frac{283}{20}\cdot\frac{1.34 - 1.07}{1.34 + 1} =
14.2\times\frac{0.27}{2.34} = 14.2\times 0.115 = 1.6.
\]
The module pumps about $1.6\,\si{\watt}$ of heat for each watt of electrical
input across a modest lift, but the coefficient collapses toward zero as
$\Delta T$ approaches the $\Delta T_{\max}$ of the earlier example, where the
module runs flat out just to hold the gradient with no cooling left over. The
Peltier cooler is therefore chosen for small lifts and modest loads, a laser
diode held at temperature or a detector cooled a few tens of kelvin, and it
loses badly to a vapour-compression cycle for any large heat load, the trade
$ZT$ quantifies.

\paragraph{LED wall-plug efficiency against luminous efficacy.}
The white-LED case study tracked two different efficiencies that are easy to
conflate, and a number separates them. The wall-plug efficiency is a pure
power ratio, optical watts out over electrical watts in, and for the case
study's chain it was $0.85\times 0.82\times 0.70\times 0.80 = 0.39$. The
luminous efficacy is lumens out over electrical watts in, so it folds in the
eye's weighting through the luminous efficacy of radiation, near
$320\,\si{\lumen\per\watt}$ for the warm-white spectrum. The two are linked
by that spectral weighting,
\[
\text{efficacy} = \eta_{\text{WPE}}\times(\text{LER})
= 0.39\times 320 = 125\,\si{\lumen\per\watt},
\]
so a $10\,\si{\watt}$ luminaire delivers $10\times 0.39 = 3.9\,\si{\watt}$
of optical power but $10\times 125 = 1250\,\si{\lumen}$. A designer who
reports only wall-plug efficiency hides the spectral penalty, and one who
reports only efficacy hides whether the loss is electrical or spectral; the
honest specification carries both, because the remedies differ, better
extraction lifts the wall-plug term while a narrower-band red phosphor lifts
the luminous efficacy of radiation.

\paragraph{Cutoff wavelength of an infrared detector.}
A photon detector responds only to photons above its band gap, so its long-
wavelength cutoff is $\lambda_c = 1.24/E_g[\si{\electronvolt}]\,\si{\micro
\meter}$. To image the $9.7\,\si{\micro\meter}$ room-temperature blackbody
peak the microbolometer vignette named, a photon detector needs a gap below
$E_g = 1.24/9.7 = 0.128\,\si{\electronvolt}$. Mercury-cadmium-telluride,
$\mathrm{Hg_{1-x}Cd_xTe}$, tunes its gap with composition $x$ from zero
(semimetal) up through the infrared, and $x \approx 0.2$ places the gap near
$0.1\,\si{\electronvolt}$, cutoff near $12\,\si{\micro\meter}$, the standard
long-wave infrared material. The cost is dark current: with $k_BT =
0.026\,\si{\electronvolt}$ at room temperature the ratio $E_g/2k_BT =
0.128/0.052 = 2.5$ is small, so $n_i \sim \exp(-2.5)$ leaves an enormous
thermal-carrier density and a dark current that buries the signal, which is
why such detectors are cooled to $77\,\si{\kelvin}$ where $E_g/2k_BT$ climbs
to about $10$ and the dark current falls by orders of magnitude. The
worked cutoff is the reason the chapter's thermal-imaging choice is between
a cooled narrow-gap photon detector and the uncooled microbolometer.

\paragraph{Thermal de~Broglie wavelength and the onset of degeneracy.}
The intrinsic-carrier formula treated the carriers with Boltzmann
statistics, valid only while they are dilute enough not to overlap
quantum-mechanically, and the \engterm{thermal de~Broglie wavelength} sets
the boundary. A carrier at temperature $T$ has a thermal wavelength
\[
\Lambda = \frac{h}{\sqrt{2\pi m^{*}k_BT}},
\]
and the Boltzmann description holds while the carrier spacing $n^{-1/3}$
stays much larger than $\Lambda$. For silicon electrons ($m^{*} =
0.26\,m_e$) at $300\,\si{\kelvin}$,
\[
\Lambda = \frac{6.63\times 10^{-34}}
               {\sqrt{2\pi(0.26\times 9.11\times 10^{-31})(1.38\times
                10^{-23})(300)}}
        = 8.7\times 10^{-9}\,\si{\meter} = 8.7\,\si{\nano\meter},
\]
so the carriers overlap and become degenerate when their density approaches
$\Lambda^{-3} = (8.7\times 10^{-9})^{-3} = 1.5\times 10^{24}\,\si{\per\meter
\cubed} = 1.5\times 10^{18}\,\si{\per\centi\meter\cubed}$. A doping above
about $10^{18}\,\si{\per\centi\meter\cubed} $ therefore pushes silicon into
the degenerate regime where the Fermi level enters the band, the $+1$ in the
Fermi-Dirac denominator can no longer be dropped, and the effective density
of states $N_c$, $N_v$ of the intrinsic formula ceases to be constant. The
example draws the exact line the earlier derivation flagged: the Boltzmann
algebra of the chapter is a non-degenerate approximation, and heavy doping
crosses out of it when the carrier spacing shrinks to the thermal
wavelength.

\paragraph{Concentration gain in the open-circuit voltage.}
A concentrator photovoltaic system focuses sunlight by a factor $C$ onto a
small cell, and the voltage benefit is worth carrying because it partly pays
for the optics. Concentrating by $C$ raises the short-circuit current by $C$
while the saturation current $J_0$ is unchanged, so the open-circuit voltage
rises logarithmically,
\[
\Delta V_{oc} = \frac{k_BT}{e}\ln C.
\]
At $C = 500\,\text{suns}$ and room temperature, $\Delta V_{oc} = 0.0259\,
\ln 500 = 0.0259\times 6.21 = 0.161\,\si{\volt}$, so a cell that gives
$0.74\,\si{\volt}$ at one sun gives about $0.90\,\si{\volt}$ under $500$-fold
concentration, and the efficiency rises with it because the extra voltage is
free power. The concentrator also lets the expensive high-efficiency III-V
cell be shrunk by the factor $C$, trading cell area for cheap mirror or lens
area, which is the economic case for concentrator systems. The limits are
that the cell heats under the focused flux, and the temperature coefficient
of the earlier example erodes the gain, and that concentration collects only
the direct beam and not the diffuse sky, so a concentrator needs tracking
and a clear-sky site to beat a flat silicon panel.

\paragraph{Diode reverse-recovery time and the stored charge.}
The diode section treated the junction in steady state, but a diode switched
from forward conduction to reverse blocking does not turn off at once, and the
delay caps its use as a fast switch. Forward current $I_F$ leaves a store of
minority charge in the neutral regions of magnitude $Q = I_F\tau$, with $\tau$
the minority-carrier lifetime, and the junction cannot support reverse voltage
until that charge is removed. When the circuit reverses the current to $I_R$, the
storage time to sweep the charge out follows from the charge balance
\[
t_s \approx \tau\,\ln\!\left(1 + \frac{I_F}{I_R}\right),
\]
the reverse phase of Figure~\ref{fig:vol04ch12:reverse-recovery}. For a
rectifier with $\tau = 1\,\si{\micro\second}$ switching from $I_F = 1\,\si{
\ampere}$ forward to $I_R = 2\,\si{\ampere}$ reverse, $t_s = 10^{-6}\ln(1.5) =
0.41\,\si{\micro\second}$, and the recovery tail roughly doubles it to a
reverse-recovery time near $t_{rr} \approx 0.8\,\si{\micro\second}$. During that
time the diode conducts backward, dumping the stored charge $Q = I_F\tau =
1\,\si{\micro\coulomb}$ as a reverse-current spike that heats the diode and
injects switching noise. The two cures read straight off the formula: shorten the
lifetime by gold-doping or electron-irradiating the silicon, which trades a higher
forward drop for a faster recovery; or use a Schottky diode, a metal-semiconductor
junction whose forward current is majority-carrier and stores no minority charge
at all, so $t_{rr}$ collapses to the junction-capacitance charging time and the
diode switches in nanoseconds. The stored-charge delay is why the humble rectifier
becomes the limiting element in a fast switching supply and why the diode's
lifetime, invisible in the steady-state Shockley law, sets its switching speed.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-reverse-recovery.tex}

\paragraph{Photodiode junction capacitance and the receiver bandwidth.}
The receiver examples specified a small photodiode without saying why area is the
enemy of speed, and the junction capacitance makes the trade a number. A
reverse-biased junction of area $A$ and depletion width $W$ is a parallel-plate
capacitor of capacitance $C_j = \varepsilon A/W$, and feeding a transimpedance
amplifier of input resistance $R$ this capacitance sets a first-order bandwidth
$f_{3\text{dB}} = 1/(2\pi R C_j)$. Take a silicon PIN photodiode of diameter
$200\,\si{\micro\meter}$, area $A = \pi(10^{-4})^{2} = 3.1\times 10^{-8}\,\si{
\meter\squared}$, with an intrinsic layer $W = 10\,\si{\micro\meter}$ wide set by
the reverse bias. The capacitance is
\[
C_j = \frac{\varepsilon A}{W}
    = \frac{(11.7\times 8.85\times 10^{-12})(3.1\times 10^{-8})}
           {10\times 10^{-6}}
    = 3.2\times 10^{-13}\,\si{\farad}
    = 0.32\,\si{\pico\farad}.
\]
Into a $10\,\si{\kilo\ohm}$ transimpedance the bandwidth is $f_{3\text{dB}} =
1/(2\pi\times 10^{4}\times 3.2\times 10^{-13}) = 50\,\si{\mega\hertz}$, enough for
the $100\,\si{\mega\bit\per\second}$ link of the design exercise only if the gain
is dropped or the diode shrunk. Halving the diameter quarters the area and the
capacitance, quadrupling the bandwidth to $200\,\si{\mega\hertz}$, at the cost of
a smaller catch for the incoming beam; widening the intrinsic layer with more
reverse bias also cuts the capacitance but lengthens the carrier transit time,
which sets a competing transit-time bandwidth $\sim v_{sat}/W$. The fast receiver
therefore lands on an optimum width where the capacitance-limited and
transit-limited bandwidths cross, the design point that fixes both the diode size
and its bias, and it is the same depletion electrostatics the varactor sold as a
tuning element, here read as the enemy of receiver speed.

\paragraph{Optical-link power budget and the amplifier spacing.}
The laser and photodiode examples fixed a launch power and a receiver
sensitivity; the link budget in between decides how far a span may run before it
needs an amplifier. Power in an optical link is bookkept in decibels relative to a
milliwatt, and every loss term subtracts (Figure~\ref{fig:vol04ch12:link-budget}).
A distributed-feedback laser launches $0\,\text{dBm}$ ($1\,\si{\milli\watt}$)
into single-mode fibre whose attenuation at the $1.55\,\si{\micro\meter}$
low-loss window is $\alpha = 0.25\,\si{\decibel\per\kilo\meter}$. Two connectors
at $0.5\,\text{dB}$ each and a pair of fusion splices at $0.1\,\text{dB}$ each
take $1.2\,\text{dB}$ of fixed loss, and the receiver sensitivity for the target
error rate is $-31\,\text{dBm}$. Reserving a $4\,\text{dB}$ margin for ageing and
bends, the fibre may spend
\[
L_{\text{fibre}} = \frac{P_{\text{launch}} - P_{\text{sens}}
  - L_{\text{fixed}} - M}{\alpha}
= \frac{0 - (-31) - 1.2 - 4}{0.25}
= \frac{25.8}{0.25}
= 103\,\si{\kilo\meter},
\]
so an unamplified span reaches about $100\,\si{\kilo\meter}$ at this bit rate,
the reach the figure closes. The budget shows the two levers a network designer
pulls: a lower-loss fibre or a longer-wavelength window stretches the reach
linearly through $\alpha$, while a more sensitive receiver, an avalanche
photodiode or coherent detection, adds decibels directly to the numerator. Past
the amplifier-free reach the span is broken by an erbium-doped fibre amplifier
every eighty to a hundred kilometres, and the same budget run per hop sets how
many amplifiers a transoceanic cable carries. The link budget is the photon-count
argument of the failure section written in decibels: each term is a factor by
which the arriving photon flux is reduced, and the span closes only while the
received flux stays above the shot-noise floor of the receiver.

\paragraph{Internal quantum efficiency from the recombination rates.}
The LED and laser sections quoted an internal quantum efficiency without deriving
it, and the ABC recombination model builds it from three competing rates that
also explain the droop. In an active region carrying carrier density $n$, three
processes remove carriers: defect (Shockley-Read-Hall) recombination at rate
$An$, radiative recombination at rate $Bn^{2}$, and Auger recombination at rate
$Cn^{3}$. Only the radiative term makes light, so the internal quantum efficiency
is the radiative rate over the total,
\[
\eta_{\text{IQE}} = \frac{Bn^{2}}{An + Bn^{2} + Cn^{3}}.
\]
Take a good InGaN emitter with $A = 10^{7}\,\si{\per\second}$, a radiative
coefficient $B = 10^{-11}\,\text{cm}^{3}\,\si{\per\second}$, and an Auger
coefficient $C = 10^{-30}\,\text{cm}^{6}\,\si{\per\second}$, evaluated at a
moderate $n = 10^{18}\,\si{\per\centi\meter\cubed}$.
The three rates are $An = 10^{7}\times 10^{18} = 10^{25}$, $Bn^{2} = 10^{-11}
\times 10^{36} = 10^{25}$, and $Cn^{3} = 10^{-30}\times 10^{54} = 10^{24}$, all in
$\si{\per\centi\meter\cubed\per\second}$, so
\[
\eta_{\text{IQE}} = \frac{10^{25}}{10^{25} + 10^{25} + 10^{24}}
= \frac{1.0}{2.1} = 0.48.
\]
At this carrier density the defect and radiative terms are comparable and the
Auger term is still small. Raising the drive to $n = 10^{19}\,\si{\per\centi
\meter\cubed}$ multiplies $An$ by ten, $Bn^{2}$ by a hundred, and $Cn^{3}$ by a
thousand, so the Auger term climbs to $10^{27}$ and overtakes the others, dropping
the efficiency and reproducing the droop of the white-LED case study: the
efficiency peaks where the radiative $Bn^{2}$ has outrun the defect $An$ but the
cubic Auger $Cn^{3}$ has not yet caught up. The model names the three levers,
cleaner material to cut $A$, stronger overlap to raise $B$, and lower carrier
density at fixed output to suppress $C$, and it is the arithmetic behind every
statement in the chapter about internal quantum efficiency.

\paragraph{Superconducting energy gap and the operating temperature.}
The Josephson vignette assumed a superconductor without saying what sets its
critical temperature or why the junction must be cooled, and the BCS energy gap
supplies both. Below the critical temperature $T_c$ the electrons bind into
Cooper pairs, and breaking a pair costs an energy $2\Delta$, where the
zero-temperature gap is fixed by the BCS relation to the critical temperature,
\[
2\Delta(0) \approx 3.5\,k_B T_c.
\]
For niobium, the workhorse of Josephson electronics with $T_c = 9.3\,\si{\kelvin}$,
the gap is $2\Delta = 3.5\times(1.38\times 10^{-23})(9.3) = 4.5\times 10^{-22}\,\si{
\joule} = 2.8\,\si{\milli\electronvolt}$, a tiny energy that sets the whole
operating regime. Two consequences follow. The junction must run far below $T_c$,
typically at $4.2\,\si{\kelvin}$ in liquid helium, because thermal energy $k_B T$
must stay well under the gap or thermally broken pairs (quasiparticles) short out
the supercurrent; at $4.2\,\si{\kelvin}$, $k_B T = 0.36\,\si{\milli\electronvolt}$
is about an eighth of the gap, cold enough. And the gap sets the maximum Josephson
frequency and the switching speed: a junction can respond up to the gap frequency
$2\Delta/h = (4.5\times 10^{-22})/(6.63\times 10^{-34}) = 6.8\times 10^{11}\,\si{
\hertz}$, near a terahertz, which is why superconducting logic is fast. The gap is
the superconductor's band gap, the energy an excitation costs, and it plays the
same role for the Josephson junction that the semiconductor band gap plays for
the diode: it fixes the operating temperature, the characteristic energy, and the
frequency ceiling, and the search for higher-$T_c$ materials is a search to lift
that gap into a regime a cheaper cooler can reach.

\paragraph{Electro-absorption modulator extinction and drive voltage.}
The modulator vignette claimed ten to fifteen decibels of extinction; the
Beer-Lambert law converts the field-driven absorption change into that number and
fixes the drive voltage. The transmission through a modulator of length $L$ at
absorption coefficient $\alpha$ is $T = \exp(-\alpha L)$, so the extinction ratio
between the biased off state ($\alpha_{\text{off}}$) and the unbiased on state
($\alpha_{\text{on}}$) is
\[
\text{ER} = 10\log_{10}\frac{T_{\text{on}}}{T_{\text{off}}}
          = 10\log_{10}\,e\,(\alpha_{\text{off}} - \alpha_{\text{on}})L
          = 4.34\,(\alpha_{\text{off}} - \alpha_{\text{on}})L.
\]
For the edge shift of Figure~\ref{fig:vol04ch12:eam-absorption} the operating
wavelength sees $\alpha_{\text{on}} = 0.005\,\si{\per\micro\meter}$ unbiased and
$\alpha_{\text{off}} = 0.15\,\si{\per\micro\meter}$ under bias, so over a
$100\,\si{\micro\meter}$ waveguide
\[
\text{ER} = 4.34\,(0.15 - 0.005)(100) = 4.34\times 14.5 = 63,
\]
which is optimistic; a real device with mode overlap and a graded edge reaches
about $10$ to $15\,\text{dB}$, but the arithmetic shows the levers. Insertion loss
is the on-state term, $\alpha_{\text{on}}L\times 4.34 = 0.005\times 100\times
4.34 = 2.2\,\text{dB}$, the price paid for placing the operating point close
enough to the edge for a strong off state. The drive voltage is whatever reverse
bias shifts the edge far enough to raise $\alpha_{\text{off}}$ to the design
value, typically one to two volts across an intrinsic region a few hundred
nanometres thick, giving fields of several megavolts per metre through the
quadratic Stark shift of the mastery box. The trade is explicit: a longer
modulator raises the extinction ratio linearly but raises the insertion loss and
the capacitance with it, so the length is chosen to reach the required extinction
at the lowest insertion loss the link can spend.

\paragraph{X-ray attenuation and the half-value layer.}
The X-ray subsection set the tube spectrum; the Beer-Lambert attenuation of that
beam in tissue and shielding is the number a radiographer and a shielding designer
both carry. A monoenergetic X-ray beam of intensity $I_0$ is attenuated through a
thickness $x$ as $I = I_0\exp(-\mu x)$, with $\mu$ the linear attenuation
coefficient set by the photoelectric and Compton cross sections at that photon
energy. The convenient measure is the half-value layer, the thickness that halves
the intensity,
\[
\text{HVL} = \frac{\ln 2}{\mu} = \frac{0.693}{\mu}.
\]
For $60\,\si{\kilo\electronvolt}$ photons in soft tissue $\mu \approx 0.2\,\si{
\per\centi\meter}$, so $\text{HVL} = 0.693/0.2 = 3.5\,\si{\centi\meter}$: a chest
some $20\,\si{\centi\meter}$ thick attenuates the beam by $\exp(-0.2\times 20) =
\exp(-4) = 0.018$, near six half-value layers, which is why the transmitted
image carries the shadow of the denser bone against the softer tissue. For
shielding, lead at the same energy has $\mu \approx 60\,\si{\per\centi\meter}$
and a half-value layer of about $0.1\,\si{\milli\meter}$, so a fraction of a
millimetre of lead in an apron cuts the scattered dose by many halvings. The
strong energy dependence, $\mu$ falling as the beam hardens, is why the tube
voltage sets penetration: raising it from $60$ to $120\,\si{\kilo\volt}$ lowers
$\mu$ in tissue and lets the beam through a thicker patient, at the cost of the
softer contrast between bone and tissue that the section's dose-against-information
budget must weigh. The same exponential law, read in reverse, is the design of
every X-ray shield and the reason dose falls off so steeply with the shielding
that separates a worker from the source.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-link-budget.tex}

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-deadtime-saturation.tex}

\section{Failure: quantum-noise floors and the limits of classical design}\pathtag{core}

Quantum mechanics sets fundamental noise floors that no amount of
classical engineering can suppress. Three are operationally important
in modern engineering.

\engterm{Shot noise} arises from the discreteness of charge: a current
$I$ flowing through a junction shows fluctuations of variance
$\sigma_{I}^{2} = 2eI\Delta f$ over a measurement bandwidth $\Delta
f$, where $e$ is the electron charge. A photodiode receiving an
average optical power $P$ at wavelength $\lambda$ has photocurrent
$I = \eta eP\lambda/(hc)$ (with $\eta$ the quantum efficiency) and
shot noise of variance $2eI\Delta f$. The signal-to-noise ratio is
$I/\sqrt{2eI\Delta f} = \sqrt{I/(2e\Delta f)}$, scaling with the
square root of the photon flux. Shot noise dominates the noise budget
of every optical communications receiver, every astronomical
photometer, and every photon-counting experiment. The Monte Carlo
check in \texttt{code/shot\_noise\_mc.py} confirms that the photon
count is Poisson, so its standard deviation is the square root of the
mean.

\engterm{Johnson-Nyquist noise} (thermal noise) is the voltage
fluctuation across a resistor at temperature $T$: the open-circuit
RMS voltage is
\[
V_{n} = \sqrt{4k_{B}TR\Delta f},
\]
where $R$ is the resistance and $\Delta f$ is the bandwidth, the result
Johnson measured and Nyquist explained in 1928
\cite{paper:ch05-johnson-1928, paper:ch05-nyquist-1928}. At room
temperature ($T = 290\,\si{\kelvin}$) and $1\,\si{\kilo\hertz}$
bandwidth, a $50\,\si{\ohm}$ resistor produces $V_{n} = 28\,\si{\nano
\volt}$ RMS; a $1\,\si{\mega\ohm}$ resistor produces $4\,\si{\micro
\volt}$ RMS. Johnson noise sets the noise floor of every passive
electrical measurement; suppressing it requires either cooling the
resistor (cryogenic low-noise amplifiers in radio astronomy) or
reducing $R$ (low-impedance source matching), or both. The reciprocal
of the Boltzmann constant $k_{B} = 1.380\,649\times 10^{-23}\,\si{
\joule\per\kelvin}$ is now fixed exactly in the 2019 SI, and Johnson
noise is the working metrological link between voltage and
temperature.

\engterm{Quantum-limited amplifier noise} is the unavoidable noise an
amplifier adds to a quantum signal. The standard quantum limit (SQL)
for a phase-preserving linear amplifier is half a photon per mode at
each input; this contributes a noise temperature of $T_{q} =
\hbar\omega/(2k_{B})$. At a microwave frequency of $10\,\si{\giga
\hertz}$, $T_{q} \approx 0.24\,\si{\kelvin}$; below this, modern
quantum-limited parametric amplifiers operate at $T \sim T_{q}$. At
optical frequencies, $T_{q}$ is much larger than any practical
ambient temperature, so optical amplifiers are typically already
quantum-limited.

\begin{mastery}
The half-photon noise floor of a phase-preserving amplifier is not a
hardware limit; it follows from the commutation relations of the field
operators. A linear amplifier that multiplies a coherent input by gain
$G$ must add at least $(G-1)/2$ noise photons per mode at the output to
keep the amplified state physical, which at large gain is half a photon
referred to the input. The escape is to give up phase information: a
phase-sensitive amplifier, which amplifies one quadrature and
de-amplifies the other, can in principle add no noise to the amplified
quadrature. Squeezed-light interferometry uses exactly this trade to
push gravitational-wave detectors below the standard quantum limit in a
chosen quadrature, at the cost of added noise in the conjugate one. The
engineering content is that the quantum floor can be moved between
observables but not removed.
\end{mastery}

\subsection{Flicker noise and the low-frequency floor}\pathtag{standard}

Shot noise and Johnson noise are white: their spectral density is flat with
frequency, so averaging longer or narrowing the band always buys a lower
floor. A third noise sits underneath them at low frequency and does not
average away in the same clean fashion, and any measurement that must be
slow meets it. \engterm{Flicker noise}, or $1/f$ noise, has a spectral
density that rises toward low frequency roughly as the inverse of frequency,
\[
S(f) \approx \frac{K}{f},
\]
so its power in a decade of bandwidth is the same in every decade, from
kilohertz down to the microhertz of a day-long drift. In a semiconductor
device it comes from carriers trapped and released by defect states in the
oxide and at interfaces, each trap switching the local conduction on a
characteristic time, and the sum over a broad distribution of trap times
produces the $1/f$ shape. The engineering consequence is a corner frequency,
the point where the falling $1/f$ density crosses the flat white floor: above
the corner the measurement is white-noise limited and averaging helps below
the corner it is flicker limited and slower is not quieter. A low-noise
amplifier is specified by both its white floor and its $1/f$ corner, and the
standard defence against flicker noise is to move the signal above the
corner by modulation, chopping or lock-in detection that carries a slow or
DC measurement up to a frequency where only white noise remains and then
demodulating it back down. The MRI, the photodiode receiver, and the
thermal imager all place their signals above the flicker corner by exactly
this move, which is why a lock-in amplifier is the default instrument for a
small slow signal.

\subsection{The high-temperature floor and the band gap}\pathtag{standard}

A different floor bounds the semiconductor devices of the chapter from
above in temperature rather than below in signal. Every junction relies on
the doping setting the carrier density, but the intrinsic carriers climb
with temperature as $n_i \sim \exp(-E_g/2k_BT)$, and once they overtake the
dopants the junction loses the asymmetry that makes it a diode. The
band-gap itself shrinks as the device heats, the Varshni narrowing of
Figure~\ref{fig:vol04ch12:varshni-gap}, which steepens the intrinsic climb
because the exponent carries a falling gap over a rising temperature. For
silicon the crossover for a moderately doped junction sits near
$600\,\si{\kelvin}$, which is why ordinary silicon is rated to junction
temperatures around $150\,{}^{\circ}\mathrm{C}$ with margin and why the
wide-gap materials silicon carbide and gallium nitride, whose larger $E_g$
pushes the crossover hundreds of degrees higher, are chosen for hot or
high-voltage power electronics. This high-temperature floor is the
mirror of the quantum-noise floors: both are set by the same
Boltzmann-and-band-gap statistics, one bounding how faint a signal a cooled
detector can see and the other bounding how hot a junction can run before
its own thermal carriers erase its design.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-varshni-gap.tex}

The quantum-noise floors are not the result of imperfect engineering;
they are the result of nature's quantum-mechanical structure. An
engineering design that requires signal-to-noise below the quantum
floor is impossible; the design must either reduce the noise
bandwidth, increase the signal power, or accept the quantum limit.

\begin{principle}[The quantum floor is the engineering floor]
Below the quantum-noise floor of the relevant signal (shot noise for
classical light, Johnson noise for thermal-electric measurements,
the standard quantum limit for amplification of coherent signals), no
classical engineering can recover the signal. The discipline is to
identify which noise floor governs the application and design the
system to operate within a factor of a few of that floor; designing
for less is over-engineering, designing for more is impossible.
\end{principle}

\begin{estimation}
\textbf{Question.} A consumer-grade CMOS image sensor with $1\,\si{
\micro\meter}$ pixel pitch and $2000\,e^{-}$ full-well capacity
images a scene at $1\,\si{\lux}$ illuminance with a $30\,\si{\milli
\second}$ exposure. Estimate the per-pixel photon shot noise in
electrons, and the achievable signal-to-noise ratio at full well.

\textbf{Estimate, before reading on.} State a factor-of-three-
confidence answer before reading the working.

\textbf{Working.} At $1\,\si{\lux}$, the photometric illuminance
corresponds to about $1.4\times 10^{-3}\,\si{\watt\per\meter
\squared}$ at the peak photopic wavelength. The pixel area is $10^{-12}\,
\si{\meter\squared}$. Photons per pixel per second: $1.4\times 10^{-3}
\times 10^{-12} / (h\nu) \approx 3.9\times 10^{3}\,\si{\per\second}$
at $\lambda = 550\,\si{\nano\meter}$ ($h\nu = 3.6\times 10^{-19}\,\si{
\joule}$). In a $30\,\si{\milli\second}$ exposure, $\sim 120$
photons fall on each pixel; with $\sim 50\,\%$ quantum efficiency and
$\sim 10\,\%$ lens transmission and obstruction at $f/2$, the pixel
collects only $\sim 6$ photoelectrons. This is far below the $2000\,e^{-}$
full well: the pixel does not saturate, and its signal is set by the few
photons it catches. Shot noise on $6$ electrons is $\sqrt{6}\approx 2.4$
electrons RMS, so the SNR is $6/\sqrt{6} = \sqrt{6} \approx 2.4$, about
$8\,\si{\decibel}$.

\textbf{Postmortem.} The estimate overturns the natural guess that
$1\,\si{\lux}$, an ordinary dim-interior level, would flood a pixel:
a one-micrometre pixel in a $30\,\si{\milli\second}$ frame catches only
a handful of photoelectrons, far short of its well, and its SNR is the
shot-noise-limited $\sqrt{N}\approx 2.4$ of that tiny count. Saturation
needs a far brighter scene, of order a few hundred lux; only there does
the pixel fill its $2000\,e^{-}$ well, where the shot noise is
$\sqrt{2000}\approx 45$ electrons for an SNR near $44$, or
$33\,\si{\decibel}$. The small pixel buys resolution and pays in light:
in dim scenes it is photon-starved, read-noise of $\sim 1\,e^{-}$
competes with the small signal, and the sensor lives on quantum
efficiency and exposure time. The numerical specifications are
representative of smartphone cameras current as of 2025.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Characterise the I-V curve of a diode]
Measure the current-voltage characteristic of a small-signal silicon
diode and identify the regions where the Shockley diode equation
holds and where it deviates.

\textbf{Materials.}
A small-signal silicon diode (1N4148 is the standard inexpensive
choice); a bench DC supply with current limit ($0$ to $5\,\si{\volt}$,
$10\,\si{\milli\ampere}$ limit); a multimeter capable of resolving
$0.1\,\si{\micro\ampere}$ current; a series resistor of $100\,\si{
\ohm}$; alligator clips; a logarithmic-scale plotting program.

\textbf{Procedure.}
\begin{enumerate}
  \item Wire the diode in series with the $100\,\si{\ohm}$ resistor
    and the bench supply; place the multimeter in series for current
    measurement.
  \item Sweep the supply voltage from $0$ to $1.0\,\si{\volt}$ in
    $25$-$\si{\milli\volt}$ steps; record the current at each step
    after waiting for thermal equilibrium ($\sim 5\,\si{\second}$).
  \item In the regime $V < 0.4\,\si{\volt}$, the current is below the
    multimeter's noise floor; flag these points as upper bounds.
  \item In the regime $0.4 < V < 0.7\,\si{\volt}$, the current
    increases exponentially; record carefully.
  \item In the regime $V > 0.7\,\si{\volt}$, series resistance of
    the diode (and the bench resistor) begins to dominate; the
    current departs from the exponential.
  \item Plot $\ln I$ versus $V$. The exponential regime appears as a
    straight line with slope $1/V_{T} = 1/(0.026\,\si{\volt})$ at
    room temperature.
  \item Extract the saturation current $I_{s}$ from the intercept.
\end{enumerate}

\textbf{Report.} A one-page report with the schematic, the
log-current-versus-voltage plot, the fitted slope and intercept, and
identification of the exponential, ohmic-series-resistance, and
saturation regions. Expected: the slope corresponds to a thermal
voltage within $\pm 10\,\%$ of $26\,\si{\milli\volt}$ at room
temperature; the saturation current is in the range $10^{-9}$ to
$10^{-13}\,\si{\ampere}$ for a small-signal diode.
Figure~\ref{fig:vol04ch12:diode-iv-semilog} shows the three regions a
clean sweep reveals on a semilog plot. A representative sweep is in
\texttt{data/diode\_iv\_sweep.csv}, and \texttt{code/diode\_iv\_fit.py}
performs the straight-line fit over the exponential region and recovers
$I_{s}$ and the thermal voltage; run against the sample data it returns
$26\,\si{\milli\volt}$ and $I_{s} \approx 1\,\si{\pico\ampere}$.

\input{volumes/04-energy/ch12-quantum-effects-engineers-use/figures/fig-diode-iv-semilog.tex}
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A photon has wavelength $\lambda = 405\,\si{\nano\meter}$ (blue
laser). Compute its energy in electronvolts and joules, its momentum,
and its frequency.
\paragraph{Solution sketch.} Working rule: $E = 1.24/0.405 = 3.06\,\si{
\electronvolt}$, equal to $3.06 \times 1.602\times 10^{-19} = 4.90\times
10^{-19}\,\si{\joule}$. Momentum $p = h/\lambda = 6.63\times 10^{-34}/
4.05\times 10^{-7} = 1.64\times 10^{-27}\,\si{\kilogram\meter\per
\second}$. Frequency $\nu = c/\lambda = 3\times 10^{8}/4.05\times 10^{-7}
= 7.40\times 10^{14}\,\si{\hertz} = 740\,\si{\tera\hertz}$. These match
the first row of \texttt{code/photon\_energy.py}.
\end{exercise}

\begin{exercise}
A silicon photodiode receives $1\,\si{\milli\watt}$ of light at
$\lambda = 850\,\si{\nano\meter}$. Compute the maximum possible
photocurrent (at $100\,\%$ quantum efficiency) and the shot noise of
the photocurrent in a $1\,\si{\mega\hertz}$ bandwidth.
\paragraph{Solution sketch.} Responsivity at unit efficiency is
$e\lambda/(hc) = \lambda[\si{\micro\meter}]/1.24 = 0.85/1.24 = 0.69\,\si{
\ampere\per\watt}$, so $I = 0.69\,\si{\ampere\per\watt}\times 10^{-3}\,
\si{\watt} = 0.69\,\si{\milli\ampere}$. Shot noise current $\sqrt{2eI
\Delta f} = \sqrt{2\times 1.6\times 10^{-19}\times 6.9\times 10^{-4}
\times 10^{6}} = \sqrt{2.2\times 10^{-16}} = 1.5\times 10^{-8}\,\si{
\ampere} = 15\,\si{\nano\ampere}$. The shot-noise-limited SNR is
$0.69\,\si{\milli\ampere}/15\,\si{\nano\ampere} \approx 4.6\times 10^{4}$.
\end{exercise}

\begin{exercise}
A silicon pn junction at $T = 300\,\si{\kelvin}$ has reverse-
saturation current $I_{s} = 5\times 10^{-12}\,\si{\ampere}$. Compute
the current at $V = 0.5\,\si{\volt}$, $V = 0.6\,\si{\volt}$, and
$V = 0.7\,\si{\volt}$ forward bias.
\paragraph{Solution sketch.} With $V_{T} = 0.0259\,\si{\volt}$,
$I = I_{s}\exp(V/V_{T})$. At $0.5\,\si{\volt}$: $\exp(0.5/0.0259) =
\exp(19.3) = 2.4\times 10^{8}$, so $I = 1.2\,\si{\milli\ampere}$. At
$0.6\,\si{\volt}$: $\exp(23.2) = 1.2\times 10^{10}$, $I = 58\,\si{\milli
\ampere}$. At $0.7\,\si{\volt}$: $\exp(27.0) = 5.6\times 10^{11}$,
$I = 2.8\,\si{\ampere}$. Each $0.1\,\si{\volt}$ multiplies the current
by $\exp(0.1/0.0259) \approx 47$, close to two decades; in a real diode
series resistance caps the $0.7\,\si{\volt}$ value well below
$2.8\,\si{\ampere}$.
\end{exercise}

\begin{exercise}
Compute the Larmor frequency of a proton in a $1.5\,\si{\tesla}$ MRI
magnet, given $\gamma_{p}/(2\pi) = 42.58\,\si{\mega\hertz\per\tesla}$.
\paragraph{Solution sketch.} $f = (\gamma_{p}/2\pi)B_{0} = 42.58\,\si{
\mega\hertz\per\tesla}\times 1.5\,\si{\tesla} = 63.9\,\si{\mega\hertz}$.
At $3\,\si{\tesla}$ the same proportionality gives $127.7\,\si{\mega
\hertz}$. The RF coil must be tuned to this frequency, which is why a
$1.5\,\si{\tesla}$ and a $3\,\si{\tesla}$ scanner use different coils.
\end{exercise}

\begin{exercise}
Compute the Johnson-Nyquist noise voltage of a $10\,\si{\kilo\ohm}$
resistor at room temperature in a $1\,\si{\kilo\hertz}$ bandwidth, and
the equivalent noise current in a $1\,\si{\mega\ohm}$ resistor.
\paragraph{Solution sketch.} $V_{n} = \sqrt{4k_{B}TR\Delta f}$. With
$4k_{B}T = 1.6\times 10^{-20}\,\si{\joule}$ at $290\,\si{\kelvin}$:
$V_{n} = \sqrt{1.6\times 10^{-20}\times 10^{4}\times 10^{3}} = \sqrt{1.6
\times 10^{-13}} = 4.0\times 10^{-7}\,\si{\volt} = 0.40\,\si{\micro
\volt}$. The noise current of a resistor is $I_{n} = V_{n}/R = \sqrt{4
k_{B}T\Delta f/R}$; for $1\,\si{\mega\ohm}$, $V_{n} = \sqrt{1.6\times
10^{-20}\times 10^{6}\times 10^{3}} = 4.0\,\si{\micro\volt}$ and
$I_{n} = 4.0\,\si{\micro\volt}/10^{6}\,\si{\ohm} = 4.0\,\si{\femto
\ampere}$. Larger resistance gives more noise voltage but less noise
current.
\end{exercise}

\begin{exercise}
Compute the de~Broglie wavelength of an electron accelerated through
$100\,\si{\kilo\volt}$, neglecting relativistic correction.
Recompute with the relativistic correction and report the error of
the non-relativistic calculation.
\paragraph{Solution sketch.} Non-relativistic: $\lambda = h/\sqrt{2m_{e}
eV}$. With $eV = 1.6\times 10^{-14}\,\si{\joule}$, $\sqrt{2m_{e}eV} =
\sqrt{2\times 9.11\times 10^{-31}\times 1.6\times 10^{-14}} = 1.71\times
10^{-22}\,\si{\kilogram\meter\per\second}$, so $\lambda = 6.63\times
10^{-34}/1.71\times 10^{-22} = 3.88\,\si{\pico\meter}$. Relativistic:
$p = \sqrt{2m_{e}eV(1 + eV/2m_{e}c^{2})}$ with $m_{e}c^{2} = 511\,\si{
\kilo\electronvolt}$, so the correction factor is $\sqrt{1 + 100/1022} =
\sqrt{1.098} = 1.048$, giving $\lambda = 3.88/1.048 = 3.70\,\si{\pico
\meter}$. The non-relativistic value is high by about $4.8\,\%$, which
matters at electron-microscope accuracy.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the Shockley diode equation by setting up the drift-diffusion
current under forward bias, assuming low injection and quasi-neutral
regions. State the assumptions explicitly.
\paragraph{Solution sketch.} Assume: abrupt junction, complete
ionisation, low-level injection, no recombination in the depletion
region, and Boltzmann statistics. The applied bias $V$ raises the
minority-carrier concentration at the depletion edge by the factor
$\exp(eV/k_{B}T)$ over equilibrium (the law of the junction): $n_{p}(0)
= n_{p0}\exp(eV/k_{B}T)$. In the quasi-neutral region minority carriers
obey the steady diffusion equation $D\,d^{2}\Delta n/dx^{2} = \Delta n/
\tau$, whose solution decays exponentially with diffusion length
$L = \sqrt{D\tau}$. The diffusion current at the edge is $J = eD\,
d\Delta n/dx = (eD/L)n_{p0}[\exp(eV/k_{B}T) - 1]$. Summing the electron
and hole contributions and identifying $I_{s} = eA(D_{n}n_{p0}/L_{n} +
D_{p}p_{n0}/L_{p})$ gives $I = I_{s}[\exp(eV/k_{B}T) - 1]$.
\end{exercise}

\begin{exercise}
Derive the Johnson-Nyquist formula $V_{n} = \sqrt{4k_{B}TR\Delta f}$
by considering a resistor in thermal equilibrium connected to a
transmission line, applying equipartition to each mode of the line.
\paragraph{Solution sketch.} Connect two equal resistors $R$ by a
lossless transmission line of length $\ell$ and characteristic
impedance $R$ (matched, no reflections). The line supports standing-wave
modes spaced $\Delta\nu = c/2\ell$, so the number of modes in a band
$\Delta f$ is $2\ell\Delta f/c$. In thermal equilibrium each mode
carries energy $k_{B}T$ by equipartition (the classical Rayleigh-Jeans
limit, valid for $h\nu \ll k_{B}T$). Power flowing one way along the
line is the mode energy times mode density times speed: $P = k_{B}T
\Delta f$. In equilibrium each resistor delivers this power to the
other. Modelling the noise source as an EMF $V_{n}$ behind $R$ driving
the matched load $R$, the delivered power is $V_{n}^{2}/4R$. Setting
$V_{n}^{2}/4R = k_{B}T\Delta f$ gives $V_{n}^{2} = 4k_{B}TR\Delta f$.
\end{exercise}

\begin{exercise}
Derive the shot-noise spectral density $S_{I} = 2eI$ for a
unidirectional current of mean $I$ comprised of independent
elementary-charge transits, by treating the current as a Poisson
sum of impulses.
\paragraph{Solution sketch.} Model the current as $i(t) = \sum_{k}
e\,h(t - t_{k})$, with transit times $t_{k}$ a Poisson process of rate
$\lambda$ and $h$ a normalised pulse shape, $\int h\,dt = 1$. The mean
current is $I = e\lambda$. Carson's theorem (or the Wiener-Khinchin
result for a shot-noise process) gives the one-sided power spectral
density $S_{I}(f) = 2e^{2}\lambda|H(f)|^{2}$, where $H$ is the Fourier
transform of $h$. For pulses short compared with the measurement
bandwidth, $|H(f)|^{2} \to 1$, so $S_{I} = 2e^{2}\lambda = 2eI$. The
variance in a bandwidth $\Delta f$ is $\sigma_{I}^{2} = S_{I}\Delta f =
2eI\Delta f$. The independence of transits, the Poisson assumption, is
what makes the variance equal $e$ times the mean count, the result the
Monte Carlo in \texttt{code/shot\_noise\_mc.py} confirms.
\end{exercise}

\begin{exercise}
Derive the Shockley-Queisser limit for a single-junction solar cell
by combining the detailed-balance condition (radiative recombination
balanced by absorption) with the blackbody spectrum at the
operating temperature.
\paragraph{Solution sketch.} Treat the cell as a body that absorbs
every photon above the gap $E_{g}$ from the $5800\,\si{\kelvin}$ solar
spectrum and re-emits as a $300\,\si{\kelvin}$ blackbody whose chemical
potential equals $eV$. The net current is $J(V) = e[\,\Phi_{\text{sun}}(
E_{g}) - \Phi_{\text{emit}}(E_{g}, V)\,]$, where $\Phi$ are the photon
fluxes above the gap. The short-circuit current is the absorbed solar
photon flux above $E_{g}$; the open-circuit voltage is set when emission
balances absorption. Maximising $J(V)\cdot V$ over the band gap balances
two losses: photons below $E_{g}$ are lost (favours small gap) and
photon energy above $E_{g}$ thermalises to $E_{g}$ (favours large gap).
The maximum of the product, evaluated against the AM1.5G spectrum,
peaks near $33\,\%$ at $E_{g} \approx 1.34\,\si{\electronvolt}$, the
curve of Figure~\ref{fig:vol04ch12:shockley-queisser}.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the minimum number of photons per second that the dark-
adapted human eye can detect, given typical psychophysical
measurements. State the consequent quantum efficiency of the visual
system.
\paragraph{Reference estimate.} The classic Hecht-Shlaer-Pirenne
threshold experiment found that a flash containing roughly $50$ to
$150$ photons at the cornea is seen about half the time. Allowing for
reflection at the cornea ($\sim 4\,\%$ lost), absorption in the ocular
media ($\sim 50\,\%$ lost), and the fraction missing the rod outer
segments, only about $5$ to $14$ photons are actually absorbed by rods.
The rods themselves are near single-photon detectors, so the
inefficiency is mostly delivery, not transduction: the photoreceptor
quantum efficiency is of order tens of percent while the whole-eye
efficiency from corneal photons is only $\sim 10\,\%$. The operational
consequence is that vision at threshold is shot-noise limited; the
flicker a dark-adapted observer sees is the Poisson statistics of a
handful of photons.
\end{exercise}

\begin{exercise}
Estimate the temperature at which the intrinsic carrier concentration
of silicon equals the doping density of $10^{15}\,\si{\per\centi
\meter\cubed}$, the conventional moderate-doping level. State the
operational consequence for high-temperature semiconductor electronics.
\paragraph{Reference estimate.} Set $n_{i}(T) = \sqrt{N_{c}N_{v}}\exp(
-E_{g}/2k_{B}T) = 10^{15}\,\si{\per\centi\meter\cubed}$, with the
density of states scaling as $T^{3/2}$. Solving numerically (the
bisection in \texttt{code/intrinsic\_carrier.py}) gives $T \approx
600\,\si{\kelvin}$, about $330\,{}^{\circ}\mathrm{C}$. Above this
temperature the thermally generated intrinsic carriers exceed the
dopants, the junction loses its designed asymmetry, and diodes and
transistors stop working: leakage rises sharply and the device can run
away thermally. This is why ordinary silicon is rated to junction
temperatures around $150\,{}^{\circ}\mathrm{C}$ with margin, and why
wide-gap materials such as SiC and GaN are chosen for hot or
high-voltage power electronics, where their larger $E_{g}$ pushes the
crossover hundreds of degrees higher.
\end{exercise}

\begin{exercise}
Estimate the tunnelling-current density through a $2\,\si{\nano
\meter}$-thick SiO$_{2}$ gate oxide at $V = 1\,\si{\volt}$ bias.
State the consequent leakage current density of a $50\,\si{\nano
\meter}\times 50\,\si{\nano\meter}$ MOSFET gate.
\paragraph{Reference estimate.} Direct-tunnelling models and measured
data for a $2\,\si{\nano\meter}$ SiO$_{2}$ at $1\,\si{\volt}$ put the
current density at roughly $1$ to $10\,\si{\ampere\per\centi\meter
\squared}$ (order of magnitude; the exact value is exponentially
sensitive to thickness, as \texttt{code/tunneling\_leakage.py} shows).
Take $J \approx 1\,\si{\ampere\per\centi\meter\squared}$. The gate area
is $50\times 50 = 2500\,\si{\nano\meter\squared} = 2.5\times 10^{-11}
\,\si{\centi\meter\squared}$, so the per-gate leakage is $J\times A
\approx 2.5\times 10^{-11}\,\si{\ampere} = 25\,\si{\pico\ampere}$.
Across $10^{9}$ gates on a chip this sums to tens of milliamperes of
static leakage, the standby-power problem that drove the move to
high-$k$ dielectrics.
\end{exercise}

\begin{exercise}
Estimate the dose (in milliSieverts) absorbed by a patient during a
chest X-ray. State assumptions about beam energy, time, and tissue
penetration.
\paragraph{Reference estimate.} A posterior-anterior chest film uses a
beam of roughly $120\,\si{\kilo\volt}$ peak, a few milliampere-seconds
of tube current, and a source-to-image distance near $1.8\,\si{\meter}$.
The effective dose tabulated for this examination is about $0.02\,\si{
\milli\sievert}$, with the lateral view adding a comparable amount.
Sanity check: $0.02\,\si{\milli\sievert}$ is a few days of natural
background radiation (background is $\sim 2$ to $3\,\si{\milli\sievert}$
per year), which matches the clinical rule of thumb that a single chest
radiograph carries negligible stochastic risk. A chest CT, by contrast,
delivers $\sim 5$ to $7\,\si{\milli\sievert}$, two-plus years of
background, the comparison that governs the order against the
information gained.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a one-dimensional Poisson-Schr\"odinger solver for the
electron density in a pn junction with arbitrary doping profile.
Verify against the depletion-approximation analytical solution for
a uniform-doping abrupt junction.
\paragraph{Model-answer sketch.} Discretise position on a grid. Iterate
two coupled equations to self-consistency: Poisson, $d^{2}\phi/dx^{2} =
-\rho(\phi)/\varepsilon$ with $\rho = e(p - n + N_{D}^{+} - N_{A}^{-})$
and Boltzmann (or Fermi-Dirac) carrier densities $n,p$ as functions of
the local potential; and, in the quantum-corrected version,
Schr\"odinger for the confined states feeding back into $n$. Use a
Gummel or Newton outer loop with damping to converge $\phi$. Verify by
turning off the quantum term and comparing against the
depletion-approximation result for an abrupt junction: depletion width
$W = \sqrt{2\varepsilon V_{bi}(N_{A}+N_{D})/(eN_{A}N_{D})}$ and the
parabolic potential drop across it. Agreement to a few percent away
from the junction edges confirms the solver; the quantum correction
then shows the carrier density pushed back from the oxide interface, the
effect that reduces gate capacitance in scaled devices.
\end{exercise}

\begin{exercise}
Implement a Monte Carlo simulation of shot noise on a photodiode at
mean photon-arrival rate $\lambda$ photons per second over a $1\,\si{
\milli\second}$ measurement interval. Verify that the variance of
the photon count equals the mean (Poisson statistic) and identify
the regime in which the Gaussian approximation begins to be useful.
\paragraph{Model-answer sketch.} The reference implementation is
\texttt{code/shot\_noise\_mc.py}. Draw many trials of the count in the
window, each a Poisson deviate of mean $N = \lambda\,t$, and report the
sample mean and variance. The simulation returns variance/mean within a
percent of $1$ for every mean tested, confirming the Poisson statistic
and hence shot-noise standard deviation $\sqrt{N}$. Overlaying a normal
density of the same mean and variance, the agreement is poor for
$N \lesssim 5$ (the distribution is visibly skewed and discrete) and
good for $N \gtrsim 10$ to $20$, the practical onset of the Gaussian
approximation. The takeaway for the bench: at high photon counts shot
noise is Gaussian and adds in quadrature with other noise; at low counts
the Poisson skew matters and photon-counting statistics must be used.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a silicon-photodiode-based receiver for a $100\,\si{\mega
\bit\per\second}$ optical link at $\lambda = 850\,\si{\nano\meter}$.
Specify the photodiode area, quantum efficiency, transimpedance-
amplifier gain, and the achievable shot-noise-limited sensitivity.
\paragraph{Model-answer sketch.} Bandwidth requirement $\sim 70\,\si{
\mega\hertz}$ sets a small junction (capacitance below a few picofarad),
so a $\sim 100\,\si{\micro\meter}$-diameter device. Pick $\eta \approx
0.8$ at $850\,\si{\nano\meter}$ (responsivity $0.55\,\si{\ampere\per
\watt}$). A transimpedance gain of $\sim 10\,\si{\kilo\ohm}$ converts a
$\si{\micro\ampere}$ signal to tens of millivolts while keeping the
amplifier noise below the photocurrent shot noise. For a target
bit-error rate of $10^{-9}$ ($Q \approx 6$), the required received
power is set by demanding signal current $\geq Q$ times the total noise.
At $\Delta f = 70\,\si{\mega\hertz}$ the shot-noise floor at $1\,\si{
\micro\watt}$ received is a few tens of nanoamperes, giving a
shot-noise-limited sensitivity in the $-30$ to $-35\,\text{dBm}$
range; in practice the transimpedance resistor's Johnson noise,
not shot noise, sets the floor, which is why the gain is chosen as high
as the bandwidth allows.
\end{exercise}

\begin{exercise}
Design a small-signal silicon diode for use as a temperature sensor.
Specify the forward bias current and the temperature-dependence
sensitivity, and identify the calibration procedure that converts
forward voltage to absolute temperature.
\paragraph{Model-answer sketch.} Drive the diode at a constant forward
current (a few tens of microampere to a milliampere) from a current
source, so the measured forward voltage tracks temperature. At fixed
current the forward voltage falls by about $2\,\si{\milli\volt\per
\kelvin}$ over the ordinary range, the dominant term coming from the
temperature dependence of $I_{s}$ through the band gap, not from the
explicit $V_{T}$ factor. The relation is close to linear over a
hundred-degree span, so a two-point calibration (ice point and a known
elevated temperature) fixes the slope and intercept. For higher
accuracy, biasing at two currents and taking the difference
$\Delta V = V_{T}\ln(I_{1}/I_{2})$ gives a voltage proportional to
absolute temperature that depends only on the current ratio and
fundamental constants, the principle of the silicon bandgap temperature
reference.
\end{exercise}

\begin{exercise}
Design a single-junction solar cell architecture (substrate, doping,
junction depth, top contact, anti-reflection coating) for $\eta \ge
20\,\%$ at $1\,\text{sun}$ AM1.5G illumination. State the
manufacturing constraints and the failure modes that limit the cell
lifetime.
\paragraph{Model-answer sketch.} A monocrystalline silicon wafer
($E_{g} = 1.12\,\si{\electronvolt}$, just below the Shockley-Queisser
optimum of Figure~\ref{fig:vol04ch12:shockley-queisser}) with a
lightly doped p-type base ($\sim 10^{16}\,\si{\per\centi\meter\cubed}$)
and a shallow heavily doped n$^{+}$ emitter ($\sim 0.3\,\si{\micro
\meter}$ deep) reaches $20\,\%$ readily; high-efficiency designs add a
back surface field and passivated contacts. The junction must be shallow
so blue photons absorbed near the surface reach the field before
recombining. A textured surface plus a $\sim 75\,\si{\nano\meter}$
silicon-nitride anti-reflection and passivation layer cuts reflection
and surface recombination. The top contact is a narrow metal grid
trading shading loss against series resistance. Lifetime is limited by
encapsulant yellowing, moisture ingress corroding the metallisation,
thermal-cycling fatigue of the solder joints, and potential-induced
degradation; the cell semiconductor itself outlasts the module
packaging.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{standard}

\begin{exercise}
A photodiode measurement shows a dark current that exceeds the
specified shot-noise floor by an order of magnitude. Identify three
candidate causes (excessive bias voltage causing impact ionisation,
surface contamination at the depletion-region edge, thermal
generation at junction defects) and the supplementary measurement
that distinguishes them.
\paragraph{Solution sketch.} Sweep the reverse bias: impact ionisation
shows a sharp super-linear rise of dark current as the breakdown voltage
is approached, the other two do not. Sweep temperature: thermal
generation at defects follows the $n_{i}(T)$ or $\sqrt{n_{i}}$
activation with a near-half-gap activation energy on an Arrhenius plot;
surface leakage has a weaker, often near-ohmic temperature dependence
and tracks humidity and surface treatment. So a current-versus-bias
curve isolates impact ionisation, and a current-versus-temperature
Arrhenius plot separates bulk thermal generation (half-gap slope) from
surface contamination (shallow slope).
\end{exercise}

\begin{exercise}
A MOSFET in a digital circuit has a gate-leakage current $100\times$
the specified value at room temperature, even though the gate-oxide
thickness was measured correctly during fabrication. Identify two
candidate causes (gate-oxide defects creating localised tunnelling
paths, ionic contamination of the oxide producing space-charge
modulation) and the failure-analysis technique that distinguishes
them.
\paragraph{Solution sketch.} Localised defect paths give leakage that is
spotty across nominally identical devices, often shows abrupt
soft-breakdown steps in time (stress-induced leakage current), and
appears as discrete hot spots under emission microscopy or conductive
atomic-force microscopy. Ionic contamination (mobile $\mathrm{Na^{+}}$,
$\mathrm{K^{+}}$) shifts the threshold voltage with bias-temperature
stress and shows hysteresis and drift in capacitance-voltage curves; a
triangular-voltage-sweep or bias-temperature-stress measurement reveals
the mobile-charge density. So spatial emission microscopy points to
discrete defects, while bias-temperature-stress C-V hysteresis points to
ionic contamination.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A laser diode used in a fibre-optic transmitter fails after $10\,000$
hours of operation, less than $10\,\%$ of its rated lifetime. Identify
the candidate failure modes (catastrophic optical damage at the front
facet, dark-line defects in the active region, ohmic-contact
degradation) and the diagnostic test for each.
\paragraph{Solution sketch.} Catastrophic optical damage is sudden and
irreversible: the light output collapses in one step, and facet
inspection (optical or electron microscopy) shows a melted or cratered
facet. Dark-line defects degrade gradually: the threshold current
creeps up over hours of operation, electroluminescence imaging shows
growing dark lines or spots in the active stripe, and the failure
tracks operating current density. Ohmic-contact degradation raises the
series resistance: the forward voltage at fixed current rises over
time while the slope efficiency above threshold changes little, and a
four-point I-V measurement isolates the contact resistance. So a
light-output trace (step versus creep), facet imaging, and a forward
I-V over life separate the three.
\end{exercise}

\begin{exercise}
A photodiode's responsivity drops by $40\,\%$ after exposure to
$100\,\si{\kilo\gray}$ of ionising radiation. Identify the
mechanism (radiation-induced defects in the depletion region) and
the design strategy that would improve radiation tolerance.
\paragraph{Solution sketch.} Displacement damage and ionising dose
create deep-level traps in and near the depletion region; these act as
recombination centres that capture photogenerated carriers before they
are collected, lowering the collection efficiency and hence the
responsivity, while also raising dark current. The design responses are
to shrink the carrier path so collection beats recombination (a
thinner active region or a stronger drift field), to choose a more
radiation-tolerant material or doping, to add shielding for the expected
dose, and to anneal the device (mild heating) where the application
allows, since some defects are removed thermally. The trade is collection
volume and quantum efficiency against radiation hardness.
\end{exercise}

\begin{exercise}
A high-resolution MRI scan shows a uniform $5\,\%$ signal-intensity
gradient across the field of view, undegraded after annual
calibration. Identify the candidate causes (gradient-coil
miscalibration, B$_{0}$-field inhomogeneity, RF-coil non-uniformity)
and the supplementary scan that distinguishes them.
\paragraph{Solution sketch.} Image a uniform phantom. RF-coil
non-uniformity (the $B_{1}$ field) produces a smooth intensity shading
that changes when the coil is swapped or when a transmit-field map
($B_{1}$ map) is acquired, and it depends on flip angle; this is the
most common cause of a smooth gradient. $B_{0}$ inhomogeneity distorts
geometry and shifts frequency: a field map or a shift of the readout
gradient polarity moves the artefact, and it worsens at higher field and
near tissue interfaces. Gradient-coil miscalibration scales spatial
position: a phantom of known dimensions images with the wrong size or a
position-dependent intensity that tracks the gradient axis and reverses
with gradient polarity. So a uniform-phantom scan plus a $B_{1}$ map, a
$B_{0}$ field map, and a geometry check on a ruler phantom separate the
three.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A research group must choose between a silicon photodiode array and
a silicon avalanche photodiode for low-light fluorescence detection.
Identify the trade-off in terms of quantum efficiency, dark noise,
gain stability, and operating-voltage requirements.
\paragraph{Solution sketch.} The plain photodiode array has high
quantum efficiency, unity gain, low and stable dark current, and runs at
a few volts; its limit is that at very low light the readout-amplifier
noise dominates the small signal. The avalanche photodiode supplies
internal gain (tens to hundreds), lifting the signal above amplifier
noise, but the avalanche multiplication adds excess noise (a noise
factor that grows with gain), the gain is steeply temperature- and
bias-dependent so it needs a stabilised high-voltage supply (tens to
hundreds of volts), and its dark current is amplified too. So choose the
avalanche device when the light is faint enough that amplifier noise
would otherwise dominate and the system can hold the bias stable;
choose the plain array when the light is adequate and stability,
simplicity, and low operating voltage matter more.
\end{exercise}

\begin{exercise}
A consulting client must decide between a Si and a GaAs photovoltaic
cell for a satellite installation. Identify the trade-off in terms of
efficiency, radiation hardness, cost, and weight.
\paragraph{Solution sketch.} GaAs (and III-V multi-junction cells built
on it) sits at the Shockley-Queisser optimum, so beginning-of-life
efficiency is higher (single-junction GaAs near $28\,\%$, triple-junction
above $30\,\%$, against silicon's $\sim 20$ to $26\,\%$), and the direct
gap lets a thin absorber tolerate displacement damage better, so the
end-of-life efficiency after years of proton and electron dose is
markedly higher; the cells are also light per watt. The cost is high:
III-V epitaxy on germanium substrates is far more expensive than
silicon. On the ground that price wins and silicon dominates; in orbit,
where launch mass and end-of-life power per area dominate the cost
function, the higher efficiency and radiation hardness of GaAs justify
the premium, which is why most spacecraft fly III-V cells. The decision
turns on whether dollars per watt or watts per kilogram after
irradiation governs the system.
\end{exercise}
