\chapter{Algebra and proof discipline}
\label{vol02:ch01}

\epigraph{In order to understand mathematics, one must work
mathematics, must produce mathematics, must \emph{do} mathematics. We
must therefore distinguish, in our learning of mathematics, between the
results, the theorems and propositions on the one hand, and the
methods, the heuristics, by which they are obtained on the other.}{
George P\'{o}lya, \emph{How to Solve It}, second edition,
\cite[p.~vi]{gen:polya1945}.}

\chapmeta{Half-life: foundational. Archetypes: balance (algebraic
balance as the form of an identity), interface (notation as the
contract between writer and reader). Exercise target: 35. Project
track: analysis only. Hazard class: none. Reader path default:
standard, with sections explicitly tagged. Named cases: Mars Climate
Orbiter (1999), Air Canada Flight 143 (1983), Ariane 5 Flight 501
(1996), Hubble Space Telescope primary mirror (1990), Patriot
missile system at Dhahran (1991).}

\noindent
Algebra and proof discipline are the syntax in which the rest of
Volume II is written. The reader who finishes this chapter writes
mathematics that another engineer can audit, manipulates expressions
without losing units or signs, and can tell when a derivation has
established a result and when it has only suggested one.

\section{Algebraic identities and manipulations}\pathtag{core}

An algebraic identity is an equation that holds for every value of its
free variables in some stated domain. It is not an equation to be
solved; it is a rewriting rule that the engineer applies in either
direction. Most engineering manipulation is the disciplined use of a
small library of identities, applied in sequence, with the units and
domain carried through every step.

\subsection{The working library}

Six identities cover the bulk of routine algebraic work. We list them
in the form an engineer reads off the page, with the domain
understood as the real numbers unless stated.

\begin{enumerate}[leftmargin=*]
  \item \engterm{Distributivity}: $a(b+c) = ab + ac$ and $(a+b)c = ac
    + bc$. Distributivity is the rule that lets sums be broken across
    a multiplication. Its loss inside non-commutative or non-linear
    structures is one of the recurring surprises of later chapters.
  \item \engterm{Square of a sum}: $(a+b)^{2} = a^{2} + 2ab + b^{2}$.
    The cross-term $2ab$ is the one engineers forget; it is the term
    that turns variance propagation into a covariance equation in
    Volume I Chapter 4.
  \item \engterm{Difference of squares}: $a^{2} - b^{2} = (a-b)(a+b)$.
    This is the workhorse that rationalises denominators, simplifies
    limits before calculus arrives, and underwrites the conjugate
    trick in complex arithmetic.
  \item \engterm{Cube of a sum}: $(a+b)^{3} = a^{3} + 3a^{2}b + 3ab^{2}
    + b^{3}$. The general pattern is the binomial theorem, which we
    state below.
  \item \engterm{Geometric finite sum}:
    $1 + r + r^{2} + \cdots + r^{n-1} = \frac{r^{n} - 1}{r - 1}$
    for $r \neq 1$. The infinite version, valid only for $\abs{r} <
    1$, gives $1/(1-r)$. The convergence condition is the part most
    often dropped from a hasty derivation.
  \item \engterm{Telescoping}: a sum of differences collapses,
    $\sum_{i=1}^{n} \left(f(i+1) - f(i)\right) = f(n+1) - f(1)$. The
    pattern is the discrete cousin of the fundamental theorem of
    calculus; the reader will meet the continuous version in
    Chapter 5.
\end{enumerate}

The identities are stated for real numbers. Each holds verbatim for
complex numbers, for polynomials with real or complex coefficients,
for square matrices that commute with one another, and for any other
commutative ring. When the underlying objects do not commute, the
square-of-a-sum identity becomes $(A+B)^{2} = A^{2} + AB + BA + B^{2}$,
which is not the same as $A^{2} + 2AB + B^{2}$. The identity we
remember from school is the special case of a general fact, and
remembering only the special case is a recurring source of error in
matrix algebra.

\subsection{The binomial theorem}

The binomial theorem generalises the cube-of-a-sum.

\begin{principle}[Binomial theorem]
For any non-negative integer $n$ and any commuting elements $a$ and
$b$ of a commutative ring,
\[
(a + b)^{n} = \sum_{k=0}^{n} \binom{n}{k} a^{n-k} b^{k},
\]
where $\binom{n}{k} = \frac{n!}{k!(n-k)!}$ is the binomial
coefficient.
\end{principle}

The binomial coefficient counts the number of ways to choose $k$
objects from $n$, which is why it appears whenever the engineer asks
how many sequences of length $n$ contain exactly $k$ instances of a
chosen event. The identity has three uses the engineer will meet
repeatedly. It expands $(1+x)^{n}$ for integer $n$ in the design of
discrete probability calculations. It approximates $(1+x)^{n}$ for
small $x$ as $1 + nx + \tfrac{n(n-1)}{2}x^{2} + \cdots$, which is the
basis of small-perturbation analysis. It provides the combinatorial
identity $\binom{n}{k} = \binom{n-1}{k-1} + \binom{n-1}{k}$, which is
proved in section 1.4 by induction.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-pascal-triangle.tex}

\Cref{fig:vol02:ch01:pascal-triangle} renders the binomial
coefficients in their Pascal-triangle layout. The script
\texttt{code/binomial\_table.py} computes the same coefficients by
integer recursion and writes them to
\texttt{data/pascal\_to\_6.csv}; a reader who wants the table for
larger $n$ can rerun the script with a larger argument.

\paragraph{Worked example: expanding $(1 + x)^{6}$ by Pascal's row.}
The sixth row of \Cref{fig:vol02:ch01:pascal-triangle} reads $1, 6,
15, 20, 15, 6, 1$. Substituting $a = 1$ and $b = x$ into the
binomial theorem gives $(1 + x)^{6} = 1 + 6x + 15 x^{2} + 20 x^{3} +
15 x^{4} + 6 x^{5} + x^{6}$. For $x = 0.05$, the truncation after
the linear term, $1 + 6(0.05) = 1.30$, undershoots the true value
$(1.05)^{6} \approx 1.3401$ by about $3\%$; carrying the quadratic
term, $1 + 6(0.05) + 15(0.05)^{2} = 1.3375$, reduces the
undershoot to about $0.2\%$. The engineer who works small-$x$
expansions reads the row of Pascal's triangle as the working table
of correction terms.

\subsection{Polynomials and factorisation}

A polynomial of degree $n$ over a field is an expression
\[
p(x) = a_{n}x^{n} + a_{n-1}x^{n-1} + \cdots + a_{1}x + a_{0},
\]
with $a_{n} \neq 0$ and the coefficients $a_{0}, \ldots, a_{n}$ in the
field. The fundamental working facts are three.

A polynomial of degree $n$ has at most $n$ real roots. Over the
complex numbers, it has exactly $n$ roots when counted with
multiplicity; this is the fundamental theorem of algebra, which we
take as given here and revisit in Chapter 4 of this volume when
complex numbers receive their working introduction.

If $r$ is a root of $p(x)$, then $(x - r)$ divides $p(x)$. The factor
theorem is the working statement of this: a polynomial factors
through its roots, and a polynomial whose roots are known has its
factorisation determined up to the leading coefficient.

A monic polynomial (one with $a_{n} = 1$) of degree $n$ with roots
$r_{1}, \ldots, r_{n}$ satisfies Vieta's relations, which connect the
coefficients to the elementary symmetric polynomials of the roots:
the sum of the roots is $-a_{n-1}$, the sum of products of pairs is
$a_{n-2}$, and so on, with alternating signs. The relations are the
discrete cousin of the trace and determinant formulas the reader will
meet in Chapter 8.

\paragraph{Worked example: Vieta's relations as a design check.} A
controls engineer designs a second-order system whose characteristic
polynomial is $s^{2} + 2\zeta\omega_{n} s + \omega_{n}^{2}$, with the two
roots the closed-loop poles. Vieta's relations read the coefficients off
the roots without solving the quadratic: the sum of the two poles is
$-2\zeta\omega_{n}$ and their product is $\omega_{n}^{2}$. Suppose the
specification asks for a natural frequency $\omega_{n} =
10\,\text{rad/s}$ and a damping ratio $\zeta = 0.7$. The product of the
poles must then be $\omega_{n}^{2} = 100$ and their sum must be
$-2(0.7)(10) = -14$. A candidate pole pair $-7 \pm 7.14\mathrm{i}$ is
checked in one line: the sum is $-14$, matching, and the product is
$7^{2} + 7.14^{2} = 49 + 51 = 100$, matching. The relations turn a
two-number specification into two arithmetic checks on the poles, with
no root extraction, and they catch a transposed digit or a sign slip
immediately. Run in reverse, the same relations are how the engineer
places poles to hit a target sum and product, the pole-placement move
Volume IX builds on.

\paragraph{Worked example: Vieta on a cubic, without finding the roots.}
The relations extend to every degree, and their value grows with the
degree because root extraction grows harder while the coefficient reading
stays a one-line check. A cubic characteristic polynomial $s^{3} + a_{2}
s^{2} + a_{1} s + a_{0}$ with roots $r_{1}, r_{2}, r_{3}$ satisfies three
relations: the sum of the roots is $r_{1} + r_{2} + r_{3} = -a_{2}$, the
sum of the pairwise products is $r_{1}r_{2} + r_{1}r_{3} + r_{2}r_{3} =
a_{1}$, and the product of all three is $r_{1}r_{2}r_{3} = -a_{0}$. Take
the polynomial $s^{3} - 6 s^{2} + 11 s - 6$, which a controls engineer
might meet as the characteristic polynomial of a third-order system. The
relations predict, before any factoring, that the three poles sum to $6$,
that their pairwise products sum to $11$, and that their product is $6$. A
proposed pole set $\{1, 2, 3\}$ is checked in three lines: the sum is $1 +
2 + 3 = 6$, matching $-a_{2}$; the pairwise products sum to $1\cdot 2 +
1\cdot 3 + 2\cdot 3 = 2 + 3 + 6 = 11$, matching $a_{1}$; and the product
is $1\cdot 2\cdot 3 = 6$, matching $-a_{0}$. All three agree, so the
proposed poles are the roots, certified without solving a cubic. The same
three relations catch a transcription error at once: a proposed set
$\{1, 2, 4\}$ has the right sum-free structure at first glance but sums to
$7$, not $6$, and fails the first relation before the engineer wastes time
on the other two. The general pattern is that a degree-$n$ monic
polynomial hands the engineer $n$ independent arithmetic checks on any
proposed root set, and the checks cost far less than the root extraction
they audit.

\subsection{Completing the square}

Completing the square is the single most useful rewriting in
elementary algebra, because it converts a quadratic from the form in
which its roots are hidden to the form in which its vertex, its sign,
and its roots are all visible at once. The move rewrites
\[
x^{2} + bx = \left(x + \frac{b}{2}\right)^{2} - \frac{b^{2}}{4},
\]
which the reader verifies by expanding the right side. The general
quadratic $ax^{2} + bx + c$ with $a \neq 0$ becomes
\[
a x^{2} + b x + c
= a\left(x + \frac{b}{2a}\right)^{2} + c - \frac{b^{2}}{4a}.
\]
The vertex sits at $x = -b/(2a)$, the minimum (or maximum) value is
$c - b^{2}/(4a)$, and setting the expression to zero and solving for
$x$ recovers the quadratic formula. The discriminant $b^{2} - 4ac$ is
the quantity under the square root, and its sign decides whether the
real roots are two, one, or none.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-completing-square.tex}

\Cref{fig:vol02:ch01:completing-square} draws the move as an area
balance: the expression $x^{2} + bx$ tiles a square of side $x$ and
two strips of total area $bx$, and adding the corner square of area
$(b/2)^{2}$ completes the larger square. The same rewriting reappears
in three later settings the reader should recognise as the same
algebra. It is the step that puts a Gaussian into the standard form
$e^{-(x - \mu)^{2}/(2\sigma^{2})}$ before the integral of Chapter 6 is
evaluated. It is the step that diagonalises a quadratic form in
Chapter 10. It is the step that turns the energy of a damped
oscillator into a sum of squares in Volume III. Completing the square
is one identity wearing several uniforms.

\paragraph{Worked example: the minimum of a quadratic cost.}
A manufacturing cost per unit is modelled as $C(q) = 0.002 q^{2} - 1.6
q + 540$, with $q$ the production rate in units per hour and $C$ in
currency units. Completing the square,
\[
C(q) = 0.002\left(q^{2} - 800 q\right) + 540
= 0.002\left(q - 400\right)^{2} - 320 + 540,
\]
so the minimum cost is $C(400) = 220$ at $q = 400$ units per hour. The
completed-square form gives the optimum location and value in one
reading, with no calculus. The reader who later differentiates
$C(q)$ will find $C'(q) = 0.004 q - 1.6 = 0$ at the same $q = 400$;
the two routes agree, and completing the square is the route that does
not need Chapter 5.

\subsection{The geometric sum, in full}

The geometric finite sum earns a subsection of its own, because it is
the identity an engineer reaches for whenever a quantity is scaled by a
constant factor at each step and the running total is wanted. The finite
form is proved by the telescoping trick, which is worth writing once so
the convergence condition on the infinite form is not mistaken for an
afterthought. Let $S = 1 + r + r^{2} + \cdots + r^{n-1}$. Multiply by
$r$ to get $rS = r + r^{2} + \cdots + r^{n}$, and subtract:
\[
S - rS = 1 - r^{n},
\qquad\text{so}\qquad
S = \frac{1 - r^{n}}{1 - r}
\quad (r \neq 1).
\]
The subtraction is the telescoping of section 1.1's working library:
every intermediate power cancels between $S$ and $rS$, leaving only the
endpoints. The infinite sum is the limit of the finite one as $n \to
\infty$. The term $r^{n}$ tends to zero exactly when $\abs{r} < 1$, and
only then does the sum converge, to $1/(1 - r)$. When $\abs{r} \geq 1$
the term $r^{n}$ does not vanish and the series diverges; the closed
form $1/(1 - r)$ is then a formal expression with no numerical meaning,
and an engineer who writes it for $\abs{r} \geq 1$ has dropped the
licensing condition exactly as the cancellation rule warned against.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-geometric-series.tex}

\Cref{fig:vol02:ch01:geometric-series} plots the partial sums for two
ratios. The tail after $n$ terms is $r^{n}/(1 - r)$, which is the
finite-sum closed form subtracted from the infinite one, and it sets
how many terms an engineer keeps for a stated accuracy. The tail shrinks
by the factor $r$ per term, so a ratio near one converges slowly and a
small ratio converges fast; the reader reads the term budget off the
tail rather than guessing it.

\paragraph{Worked example: the settling of a discrete-time filter.} A
first-order digital filter updates its output by $y_{k} = a y_{k-1} +
(1 - a) u_{k}$, with $0 < a < 1$ the retention factor and $u_{k}$ the
input. Applying a unit step $u_{k} = 1$ from rest $y_{0} = 0$, the
output after $n$ steps unrolls into a geometric sum,
\[
y_{n} = (1 - a) \sum_{k=0}^{n-1} a^{k}
= (1 - a) \cdot \frac{1 - a^{n}}{1 - a}
= 1 - a^{n}.
\]
The steady state is $1$, the limit as $n \to \infty$ because $a^{n} \to
0$ for $\abs{a} < 1$, and the shortfall from steady state after $n$
steps is exactly $a^{n}$, the geometric tail. To settle within
$2\%$ of the final value the engineer solves $a^{n} \leq 0.02$, that
is, $n \geq \ln(0.02)/\ln(a)$. For $a = 0.9$ this gives $n \geq 37$
steps; for $a = 0.5$ it gives $n \geq 6$. The retention factor that
makes the filter smooth is the same factor that makes it slow, and the
geometric identity turns the trade-off into a number the designer can
quote before building the filter. The reader meets the same $a^{n}$ tail
as the transient of a first-order differential equation in Chapter 7,
where the discrete power $a^{n}$ becomes the continuous exponential
$e^{-t/\tau}$.

\subsection{Polynomial evaluation: Horner's nested form}

A polynomial is most efficiently and most accurately evaluated in
nested form. Writing
\[
p(x) = a_{n} x^{n} + a_{n-1} x^{n-1} + \cdots + a_{1} x + a_{0}
= \Bigl(\cdots\bigl((a_{n} x + a_{n-1}) x + a_{n-2}\bigr) x +
\cdots\Bigr) x + a_{0}
\]
turns the evaluation into a sweep from the highest coefficient down,
multiplying the running total by $x$ and adding the next coefficient
at each step. The nested form, Horner's method, uses $n$
multiplications and $n$ additions. The naive term-by-term form, which
recomputes each power $x^{k}$, uses on the order of $n^{2}/2$
multiplications. \Cref{fig:vol02:ch01:pascal-triangle} is the
coefficient source for one such polynomial; the operation counts for
both methods, as a function of degree, are tabulated in
\texttt{data/horner-vs-naive-ops.csv}, and the reference
implementation is \texttt{code/horner\_eval.py}.

\paragraph{Worked example: evaluating $(1+x)^{6}$ by Horner.}
With the sixth Pascal row $1, 6, 15, 20, 15, 6, 1$ as coefficients
(highest degree first), Horner evaluates $(1 + x)^{6}$ at $x = 0.05$
by the sweep $1 \to 1 \cdot 0.05 + 6 = 6.05 \to 6.05 \cdot 0.05 + 15 =
15.3025 \to \cdots$, reaching $1.3400956$ after six steps, matching
$(1.05)^{6}$ to the displayed digits. The sweep also yields the
derivative $p'(x)$ for free, by accumulating a running derivative
alongside the running value; this synthetic-division-with-derivative
form is the engine of the Newton-Horner root-finder of Chapter 16. The
nested form matters beyond speed: it avoids forming the large
intermediate powers that the naive sum then partly cancels, which is
the accuracy argument the reader meets again in the numerical-methods
chapter \cite{text:higham-numerical}.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-horner-ops.tex}

\Cref{fig:vol02:ch01:horner-ops} plots the two operation counts against
degree. The linear Horner count and the quadratic naive count separate
early: at degree ten the naive form does five times the work, at degree
twenty ten times. The separation is the visible form of the same $n$
against $n^{2}/2$ comparison the paragraph stated, and it is the reason
a library polynomial evaluator is written in nested form without
exception.

\paragraph{Worked example: Horner carrying value and derivative
together.} The nested sweep computes more than the value. Alongside the
running value $b_{k}$ that Horner accumulates, an engineer can carry a
running derivative $d_{k}$ with the twin recurrence $d_{k} = d_{k+1} x +
b_{k+1}$, initialised at $d_{n} = 0$ and $b_{n} = a_{n}$. We work the
quadratic $p(x) = 2x^{2} - 3x + 5$ at $x = 4$, whose exact value is
$32 - 12 + 5 = 25$ and whose exact derivative $p'(x) = 4x - 3$ is
$13$. The value sweep runs $b_{2} = 2$, then $b_{1} = 2 \cdot 4 + (-3)
= 5$, then $b_{0} = 5 \cdot 4 + 5 = 25$, recovering the value. The
derivative sweep runs alongside: $d_{2} = 0$, then $d_{1} = 0 \cdot 4 +
b_{2} = 2$, then $d_{0} = 2 \cdot 4 + b_{1} = 8 + 5 = 13$, recovering
the derivative. Both come out of one pass with $2n$ multiplications
total. The paired recurrence is synthetic division carried one order
further, and it is the exact update the Newton-Horner root-finder of
Chapter 16 needs at every iteration: a root step wants $p(x_{k})$ and
$p'(x_{k})$ at the current guess, and this sweep delivers both without
forming the derivative polynomial separately.

\subsection{Partial fractions: a preview of a decomposition}

A rational function, a ratio of polynomials, often needs to be
rewritten as a sum of simpler fractions before it can be integrated,
inverse-transformed, or read for its behaviour near a pole. The
decomposition of
\[
\frac{1}{(x - a)(x - b)}
= \frac{1}{a - b}\left(\frac{1}{x - a} - \frac{1}{x - b}\right),
\qquad a \neq b,
\]
is the simplest case, and the reader verifies it by combining the
right side over a common denominator. Partial fractions is an algebra
move that pays off in three later chapters: Chapter 6 uses it to
integrate rational functions; Chapter 7 uses it to invert Laplace
transforms of linear systems; Volume IX uses it to read the
time-response of a transfer function from the locations of its poles.
The full algorithm, including repeated and complex roots, is deferred
to Chapter 6, where it earns its place against a concrete integral.
The working point here is that the decomposition is an identity, valid
for every $x$ outside the poles, and that finding the constants is a
matter of clearing denominators and matching coefficients, the same
discipline as every other identity in this section.

\paragraph{Worked example: a full partial-fraction decomposition.}
The simplest case hides the two complications that make the general
algorithm worth stating: a repeated factor and a quadratic factor with
no real root. We carry one expression through that exhibits both, so
that the later chapters can cite the move rather than re-derive it. Take
\[
F(x) = \frac{2x^{2} + 3}{(x - 1)^{2}\,(x^{2} + 1)},
\]
a rational function with a double pole at $x = 1$ and a conjugate pair
of poles at $x = \pm \mathrm{i}$. The degree of the numerator (two) is
below the degree of the denominator (four), so no polynomial part is
split off first; when the numerator degree equals or exceeds the
denominator degree, the first step is polynomial long division, and the
partial-fraction work is applied to the remainder. The decomposition
ansatz attaches one term to each power of the repeated linear factor and
a linear numerator to the irreducible quadratic,
\[
\frac{2x^{2} + 3}{(x - 1)^{2}(x^{2} + 1)}
= \frac{A}{x - 1} + \frac{B}{(x - 1)^{2}}
+ \frac{Cx + D}{x^{2} + 1}.
\]
Clearing the denominator gives the polynomial identity
\[
2x^{2} + 3 = A(x - 1)(x^{2} + 1) + B(x^{2} + 1)
+ (Cx + D)(x - 1)^{2},
\]
which must hold for every $x$. Two routes find the constants. The
cover-up route substitutes the poles. At $x = 1$ the first and third
terms vanish, leaving $2(1) + 3 = B(1 + 1)$, so $B = 5/2$. At $x =
\mathrm{i}$ the first two terms vanish (their $x^{2} + 1$ factor is
zero), leaving $2(\mathrm{i})^{2} + 3 = (C\mathrm{i} + D)(\mathrm{i} -
1)^{2}$. The left side is $2(-1) + 3 = 1$. The right side uses
$(\mathrm{i} - 1)^{2} = \mathrm{i}^{2} - 2\mathrm{i} + 1 = -2\mathrm{i}$,
so $1 = (C\mathrm{i} + D)(-2\mathrm{i}) = -2 C \mathrm{i}^{2} - 2 D
\mathrm{i} = 2C - 2D\mathrm{i}$. Matching real and imaginary parts,
$2C = 1$ and $-2D = 0$, so $C = 1/2$ and $D = 0$. The coefficient $A$
follows from the remaining freedom: matching the coefficient of
$x^{3}$ on both sides gives $0 = A + C$, so $A = -1/2$. The
decomposition is
\[
F(x) = \frac{-1/2}{x - 1} + \frac{5/2}{(x - 1)^{2}}
+ \frac{x/2}{x^{2} + 1}.
\]
The check is the balance test of this section: substitute one
convenient value, say $x = 0$, into both the original and the
decomposed forms. The original gives $3/(1 \cdot 1) = 3$; the
decomposition gives $(-1/2)(-1) + (5/2)(1) + 0 = 1/2 + 5/2 = 3$. The
two agree, which is necessary but not sufficient; a second value, $x =
2$, gives the original $11/(1 \cdot 5) = 11/5$ and the decomposition
$-1/2 + 5/2 + 1/5 = 2 + 1/5 = 11/5$, and two agreements on a
degree-three numerator identity that has four free constants leave the
result as solidly checked as hand verification reaches. The three
forms of term map onto three forms of later behaviour: the simple pole
$A/(x - 1)$ integrates to a logarithm and inverts to a decaying
exponential, the double pole $B/(x - 1)^{2}$ integrates to a power and
inverts to a term that grows like $t e^{t}$, and the quadratic term
$(Cx + D)/(x^{2} + 1)$ integrates to a logarithm-plus-arctangent and
inverts to an oscillation. Chapter 6 reads the integral off this form;
Chapter 7 reads the transient off it; Volume IX reads the stability of
a control loop off the sign of the real parts of the poles. The
algebra done once here is reused, unchanged, in each.

\subsection{Worked example: the square of a sum when the factors do
not commute}

The square-of-a-sum identity carries a hidden assumption that the two
terms commute, and the assumption fails for matrices, the objects the
reader manipulates for the rest of the volume. We work a concrete
$2\times 2$ case so the missing term is a number on the page rather than
a warning in the abstract. Take
\[
A = \begin{pmatrix} 1 & 1 \\ 0 & 1 \end{pmatrix},
\qquad
B = \begin{pmatrix} 1 & 0 \\ 1 & 1 \end{pmatrix}.
\]
The school identity would predict $(A+B)^{2} = A^{2} + 2AB + B^{2}$. We
compute all the pieces and read off the discrepancy. The sum is
$A + B = \begin{psmallmatrix} 2 & 1 \\ 1 & 2 \end{psmallmatrix}$, whose
square is
\[
(A+B)^{2}
= \begin{pmatrix} 2 & 1 \\ 1 & 2 \end{pmatrix}
  \begin{pmatrix} 2 & 1 \\ 1 & 2 \end{pmatrix}
= \begin{pmatrix} 5 & 4 \\ 4 & 5 \end{pmatrix}.
\]
The two cross products are different:
\[
AB = \begin{pmatrix} 2 & 1 \\ 1 & 1 \end{pmatrix},
\qquad
BA = \begin{pmatrix} 1 & 1 \\ 1 & 2 \end{pmatrix},
\qquad
AB - BA = \begin{pmatrix} 1 & 0 \\ 0 & -1 \end{pmatrix} \neq 0.
\]
With $A^{2} = \begin{psmallmatrix} 1 & 2 \\ 0 & 1 \end{psmallmatrix}$ and
$B^{2} = \begin{psmallmatrix} 1 & 0 \\ 2 & 1 \end{psmallmatrix}$, the
correct expansion $(A+B)^{2} = A^{2} + AB + BA + B^{2}$ sums to
\[
\begin{pmatrix} 1 & 2 \\ 0 & 1 \end{pmatrix}
+ \begin{pmatrix} 2 & 1 \\ 1 & 1 \end{pmatrix}
+ \begin{pmatrix} 1 & 1 \\ 1 & 2 \end{pmatrix}
+ \begin{pmatrix} 1 & 0 \\ 2 & 1 \end{pmatrix}
= \begin{pmatrix} 5 & 4 \\ 4 & 5 \end{pmatrix},
\]
which matches the direct square. The school prediction $A^{2} + 2AB +
B^{2}$, by contrast, would replace $AB + BA$ with $2AB =
\begin{psmallmatrix} 4 & 2 \\ 2 & 2 \end{psmallmatrix}$ and give
$\begin{psmallmatrix} 7 & 4 \\ 4 & 4 \end{psmallmatrix}$, which is wrong
in every entry that the commutator $AB - BA$ touches. The quantity that
measures the failure is exactly that commutator, and the corrected
identity is $(A+B)^{2} = A^{2} + B^{2} + AB + BA = A^{2} + B^{2} + 2AB -
(AB - BA)$. The engineer who carries the commutator as a bookkeeping
term never loses the cross structure, and the reader meets the same
commutator as the obstruction to simultaneous diagonalisation in Chapter
8 and as the source of the ordering rules for rotations in Volume III.

\begin{mastery}
The identities of this section are stated for a \engterm{commutative
ring}: a set with an addition and a multiplication that are
associative and commutative, with a distributive law tying them
together, an additive identity $0$ and inverse, and a multiplicative
identity $1$. The integers $\Z$, the rationals $\Q$, the reals $\R$,
the complex numbers $\C$, polynomials over any of these, and the
integers mod $n$ are all commutative rings. A \engterm{field} is a
commutative ring in which every nonzero element has a multiplicative
inverse; $\Q$, $\R$, $\C$, and $\Z/p\Z$ for prime $p$ are fields,
while $\Z$ and $\Z/6\Z$ are not. The distinction is exactly what
licenses cancellation: in a field, $ab = ac$ with $a \neq 0$ implies
$b = c$, because $a$ has an inverse to multiply through by. In a ring
that is not a field, the cancellation can fail, as $\Z/6\Z$ shows with
$2 \cdot 3 = 2 \cdot 0$. Square matrices under matrix addition and
multiplication form a ring that is not commutative and not a field;
this is why the square-of-a-sum identity acquires its cross term
$AB + BA$ and why $AB = AC$ does not give $B = C$ unless $A$ is
invertible. The reader does not need the full theory of rings and
fields to do engineering algebra. The reader does need to know that
the identities of school algebra carry exactly the assumptions of a
field, and that the moment the working objects leave a field, each
identity must be re-checked against the weaker structure that
remains. Halmos's \emph{Naive Set Theory} \cite{text:halmos-naive-sets}
and any first course in abstract algebra make the structure precise;
the engineering payoff is the habit of asking, before cancelling,
which structure the symbols live in.
\end{mastery}

\begin{mastery}
Beneath the ring sits a simpler structure that engineers meet more
often than they name it: the \engterm{group}. A group is a set $G$ with
one operation $\ast$ that is associative, has an identity element $e$
with $e \ast g = g \ast e = g$, and gives every element $g$ an inverse
$g^{-1}$ with $g \ast g^{-1} = e$. Commutativity is not required; a
group in which $\ast$ commutes is called abelian. The integers under
addition form an abelian group; the nonzero reals under multiplication
form an abelian group; the rotations of a rigid body in three
dimensions form a non-abelian group, because rotating about $x$ then
$y$ does not equal rotating about $y$ then $x$. The non-commutativity is
not a mathematical curiosity. It is the reason a robot arm's joint
commands do not compose by simple addition, the reason the order of two
coordinate transforms matters in a graphics pipeline, and the reason
the composition of functions in section 1.5 reverses under inverse,
$(g \circ f)^{-1} = f^{-1} \circ g^{-1}$, exactly as group inverses do.
The same axioms govern the symmetries of a crystal lattice, the
permutations a sorting network applies, the modular arithmetic of a
cyclic counter, and the error-correcting codes of Volume VII, each of
which is a group with its own operation. The working lesson is the same
as the ring box: when the engineer's objects compose, the first
question is which group law they obey and whether it is abelian,
because the answer decides whether order can be rearranged. The
identities of school algebra assumed an abelian group under addition
and, away from zero, under multiplication; the moment the objects stop
commuting, the engineer is in a non-abelian group and the rearrangement
the school identity took for granted is no longer licensed.
\end{mastery}

\subsection{Worked example: a derivation that fails on a missed
identity}

A reader sets up the discharge of a capacitor through a resistor and
needs to compute the energy dissipated. The capacitor stores charge
$Q$ at voltage $V_{0}$, with $C = Q/V_{0}$. The energy stored is the
integral of $V \,\dd Q$ from $0$ to $Q$, which a calculus chapter
will derive; for now we accept the result, $E = \tfrac{1}{2} C
V_{0}^{2}$.

The reader is asked to show that the energy dissipated in the
resistor across a full discharge equals the energy that was stored.
Setting up the equation $\tfrac{1}{2}CV_{0}^{2}$ for the dissipated
quantity, the reader computes the algebra
\[
E_{R} = \int_{0}^{\infty} \frac{V(t)^{2}}{R} \,\dd t,
\]
with $V(t) = V_{0} e^{-t/RC}$. The integrand becomes $V_{0}^{2}
e^{-2t/RC}/R$. The integral evaluates (Chapter 6) to $V_{0}^{2}/R
\cdot RC/2 = \tfrac{1}{2} C V_{0}^{2}$. The two energies match.

A common error in writing this is to expand $V(t)^{2}$ as $(V_{0}
e^{-t/RC})^{2} = V_{0}^{2} e^{-t/RC}$ rather than $V_{0}^{2}
e^{-2t/RC}$. The error is an identity slip: $(e^{x})^{2} = e^{2x}$,
not $e^{x}$. The factor of two in the exponent is what makes the
algebra balance. We make the example explicit because the discipline
of carrying exponents correctly is one of the most common places a
working engineer's algebra fails.

\begin{archetype}[Balance]
An algebraic identity is a balance. Both sides of an identity carry
the same value for every legal substitution. When an engineer
manipulates an expression, the discipline is to preserve the balance
at every step: same units, same domain, same algebraic content. A
manipulation that changes the value of the expression has broken the
balance. The archetype recurs across every chapter of this volume,
in the form of conserved quantities, equations of state, and exact
identities preserved by transformation. Chapter 8 of Volume II
formalises the balance for systems of linear equations; Volume III
formalises it for forces in equilibrium; the divergence theorem of
Chapter 11 is balance for fluxes through a closed surface. Each is a
descendant of the school identity that $a^{2} - b^{2} = (a-b)(a+b)$.
\end{archetype}

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-identity-balance.tex}

\Cref{fig:vol02:ch01:identity-balance} draws the balance pictorially.
The picture is meant as a mnemonic: when the engineer's algebra has
gone wrong, the first check is whether the move just made would
survive the balance test, in the sense that substituting a numerical
value for every variable on both sides would still produce equality.

\subsection{Estimation before manipulation}

Before any algebraic step that compresses two terms into one, the
engineer estimates the magnitude of each term. The estimate guards
against the most common failure mode: dropping a term that turns out
to have been the dominant one.

\begin{estimation}
\textbf{An aluminum cantilever beam of length $L = 0.50 \,\si{\meter}$
deflects under its own weight. The deflection is approximately
$\delta \approx \frac{\rho g L^{4}}{8 E h^{2}}$, where $\rho =
2700 \,\si{\kilogram\per\meter\cubed}$ is the density, $g = 9.81
\,\si{\meter\per\second\squared}$ is the gravitational acceleration,
$E = 70 \,\si{\giga\pascal}$ is the elastic modulus, and $h = 5
\,\si{\milli\meter}$ is the thickness. Estimate $\delta$, in
millimetres, to one significant figure, before computing.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A working estimate manipulates the algebra in halves. The numerator
$\rho g L^{4}$ has units kilogram per metre cubed times metre per
second squared times metre to the fourth, which is kilograms times
metres squared per second squared, which is joules. The denominator
$E h^{2}$ has units pascal times metre squared, which is newton per
metre squared times metre squared, which is newtons. Joules over
newtons is metres. Dimensions check.

For magnitudes, $\rho g \approx 2.7 \times 10^{4}
\,\si{\newton\per\meter\cubed}$. $L^{4} = 0.25^{2} = 0.0625
\,\si{\meter\tothe{4}}$, multiply that by $\rho g$ to get about
$1.7 \times 10^{3} \,\si{\newton\per\meter}$. Divide by $E h^{2}
\approx 70 \times 10^{9} \times 25 \times 10^{-6} = 1.75 \times
10^{6} \,\si{\newton}$. The deflection is roughly $10^{-3}
\,\si{\meter}$, or one millimetre. Including the divisor $8$, the
estimate is about $0.1$ to $0.2 \,\si{\milli\meter}$.

The exact computation gives $\delta \approx 0.13
\,\si{\milli\meter}$. The estimate is within a factor of two. The
discipline of breaking the expression into magnitudes before
arithmetic is what catches an engineer who writes $L^{4}$ as $L^{2}$
or who confuses $h^{2}$ with $h^{3}$ in the denominator.
\end{estimation}

\subsection{The cancellation rule and its limits}

A familiar algebraic move is to cancel a common factor from both
sides of an equation: $ab = ac$ implies $b = c$. The move is
legitimate when $a \neq 0$. It fails when $a = 0$, because every $b$
and every $c$ then satisfy the equation. A subtler version of the
same trap appears in mod-$n$ arithmetic and in matrix arithmetic,
where the cancellation rule fails even for nonzero factors. In
$\Z/6\Z$, $2 \cdot 3 = 0 = 2 \cdot 0$, but $3 \neq 0$. In matrix
algebra, $AB = AC$ does not imply $B = C$ unless $A$ is invertible.
The school cancellation rule is the special case of a general fact,
and the special case fails in domains the engineer will routinely
work in.

The discipline is to state the assumption that licenses the
cancellation. Under what condition is the divisor nonzero, the matrix
invertible, the polynomial coprime to the modulus, the function
injective. Stating the condition is the difference between an
algebraic step and a wish.

\paragraph{Worked example: when division mod $n$ is legal, and when it is
not.} A cyclic counter, a checksum, and a rotation index all run in the
arithmetic of $\Z/n\Z$, so an engineer who solves $a x \equiv b \pmod n$
for $x$ needs to know when the division by $a$ is licensed. The rule is
that $a$ has a multiplicative inverse mod $n$ exactly when $\gcd(a, n) =
1$, and only then does the cancellation $a x \equiv a y \Rightarrow x
\equiv y$ hold. Take $n = 12$, the arithmetic of a clock face. To solve
$5 x \equiv 3 \pmod{12}$, note $\gcd(5, 12) = 1$, so $5$ is invertible;
its inverse is $5$ itself because $5 \cdot 5 = 25 \equiv 1 \pmod{12}$, and
$x \equiv 5 \cdot 3 = 15 \equiv 3 \pmod{12}$, a unique solution. Now try
$4 x \equiv 3 \pmod{12}$: here $\gcd(4, 12) = 4 \neq 1$, so $4$ has no
inverse, and the equation has no solution at all, because $4 x$ is always
a multiple of $4$ and $3$ is not. The companion equation $4 x \equiv 8
\pmod{12}$ does have solutions, but three of them, $x \equiv 2, 5, 8$,
because the shared factor collapses the cancellation: $4 \cdot 2 = 8$,
$4 \cdot 5 = 20 \equiv 8$, and $4 \cdot 8 = 32 \equiv 8$, all mod $12$.
The single condition $\gcd(a, n) = 1$ decides all three outcomes: unique
solution when the gcd is one, and either no solution or several when it
exceeds one, according to whether the gcd divides $b$. The engineer who
writes ``divide both sides by $4$'' in modular arithmetic has assumed an
inverse that a shared factor with the modulus denies, and the counter or
checksum that relies on the false step will alias silently. The safe
habit is the same as everywhere else in this section: name the condition,
here $\gcd(a, n) = 1$, before the division is written.

\subsection{Algebraic identities that the machine breaks: catastrophic
cancellation}

An identity that is exact over $\R$ can lose almost all of its accuracy
when evaluated in finite-precision arithmetic, and the loss is governed
by algebra the engineer can predict before running the code. The
mechanism is catastrophic cancellation: subtracting two nearly equal
numbers, each carrying a small relative rounding error, produces a
difference whose relative error is amplified by the ratio of the
operands to their difference.

Quantify it. Suppose two positive quantities $a$ and $b$ are each stored
with relative error at most $u$ (the unit roundoff, about $1.1 \times
10^{-16}$ for IEEE 754 double precision), so the stored values are
$\tilde a = a(1 + \delta_{a})$ and $\tilde b = b(1 + \delta_{b})$ with
$\abs{\delta_{a}}, \abs{\delta_{b}} \leq u$. The computed difference is
$\tilde a - \tilde b = (a - b) + (a \delta_{a} - b \delta_{b})$, and the
absolute error $a \delta_{a} - b \delta_{b}$ is bounded by $(a + b) u$.
The relative error of the difference is therefore bounded by
\[
\frac{\abs{(a \delta_{a} - b \delta_{b})}}{\abs{a - b}}
\leq \frac{(a + b)\, u}{\abs{a - b}}.
\]
The amplification factor $(a + b)/\abs{a - b}$ is the condition number
of the subtraction. When $a$ and $b$ agree to $k$ significant figures,
$\abs{a - b} \approx (a + b) \cdot 10^{-k}$, the factor is about
$10^{k}$, and the subtraction has surrendered $k$ of the figures the
operands carried. The algebra of the bound is elementary; its
consequence is that the engineer who writes a difference of near-equal
quantities has written a step whose accuracy the surrounding algebra
cannot recover.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-cancellation.tex}

\Cref{fig:vol02:ch01:cancellation} plots the relative error of a
worked difference against the number of agreeing figures, on a
representative double-precision evaluation. The remedy is to rewrite the
identity so that the cancelling subtraction never forms. Three rewrites
recur. The quadratic root $x_{-} = (-b + \sqrt{b^{2} - 4ac})/(2a)$,
when $b > 0$ and $b^{2} \gg 4ac$, cancels in the numerator; the Vieta
companion $x_{-} = c/(a x_{+})$ computes it through the well-conditioned
root and a division, with no subtraction (the exercise of the
calculation set works the numbers). The expression $\sqrt{x + 1} -
\sqrt{x}$ for large $x$ cancels; multiplying by the conjugate gives
$1/(\sqrt{x + 1} + \sqrt{x})$, a sum in the denominator and no
cancellation. The expression $1 - \cos x$ for small $x$ cancels;
the half-angle identity $1 - \cos x = 2 \sin^{2}(x/2)$ replaces the
difference with a square of a small but well-resolved quantity. Each
rewrite is an algebraic identity from this chapter's working library,
applied not to simplify the expression on paper but to keep its value
when the reals are replaced by the floating-point grid of section 1.6.
The numerical-methods literature treats the bound and the rewrites in
full \cite{text:higham-numerical}; the algebraic point for this chapter
is that the identity an engineer reaches for, and the identity the
machine can evaluate accurately, are sometimes different members of the
same family, and choosing between them is an algebra decision made
before the code is written.

\paragraph{Worked example: the two-pass variance and its single-pass
trap.} The sample variance of $n$ readings $x_{1}, \ldots, x_{n}$ with
mean $\bar x$ is, by definition, $s^{2} = \frac{1}{n - 1} \sum_{i}
(x_{i} - \bar x)^{2}$. The algebraic identity
\[
\sum_{i=1}^{n} (x_{i} - \bar x)^{2}
= \sum_{i=1}^{n} x_{i}^{2} - n \bar x^{2}
= \sum_{i=1}^{n} x_{i}^{2}
- \frac{1}{n}\Bigl(\sum_{i=1}^{n} x_{i}\Bigr)^{2}
\]
is exact over $\R$, and the right-hand form is tempting because it needs
only a running sum of values and a running sum of squares, a single
pass. The trap is that for readings clustered far from zero (a
temperature series around $300\,\si{\kelvin}$, say, with millikelvin
spread) the two terms $\sum x_{i}^{2}$ and $\frac{1}{n}(\sum x_{i})^{2}$
are both enormous and nearly equal, and their difference is exactly the
catastrophic cancellation analysed above; the computed variance can come
out negative, which a variance cannot be. The two-pass form, which
computes $\bar x$ first and then sums $(x_{i} - \bar x)^{2}$, forms each
summand from a difference of order the spread rather than a difference
of order the magnitude, and is well-conditioned. The two expressions
are the same identity; one is safe and one is not. The reader meets the
incremental, numerically stable update for the same quantity in the case
study of section 1.4, where induction certifies its correctness.

\section{Inequalities}\pathtag{core}

An equation is a balance. An inequality is a directed bound: a claim
that one quantity does not exceed another. Engineers use inequalities
constantly because measured reality almost never delivers exact
equality. A specification is a tolerance, a safety factor is a bound,
a worst-case analysis is an inequality, a rate limit is an
inequality. We treat the working forms of three inequalities and
note where they are about to do load-bearing work in later chapters.

\subsection{The triangle inequality}

For real numbers $a$ and $b$,
\[
\abs{a + b} \leq \abs{a} + \abs{b}.
\]
Equality holds if and only if $a$ and $b$ have the same sign or one
of them is zero. The triangle inequality is what its name says: in
the plane, the length of one side of a triangle is at most the sum
of the lengths of the other two.

\begin{proof}
We argue by cases. If $a + b \geq 0$, then $\abs{a + b} = a + b \leq
\abs{a} + \abs{b}$ because $a \leq \abs{a}$ and $b \leq \abs{b}$. If
$a + b < 0$, then $\abs{a + b} = -(a + b) = -a + (-b) \leq \abs{a} +
\abs{b}$ for the same reason. Equality in the first case requires
$a = \abs{a}$ and $b = \abs{b}$, that is, both non-negative.
Equality in the second case requires both non-positive. The reverse
direction (equality implies same-sign) follows from inspection of
the two cases.
\end{proof}

The triangle inequality generalises to vectors in any inner-product
space, with $\abs{\cdot}$ replaced by the norm $\norm{\cdot}$. Its
direct engineering use is the bound on the magnitude of a sum of
errors: a vector quantity contaminated by independent errors of
magnitude $\epsilon_{1}, \ldots, \epsilon_{n}$ has total error of
magnitude at most $\epsilon_{1} + \cdots + \epsilon_{n}$. The bound
is loose when the errors partially cancel; the engineer learns to
distinguish the worst-case bound from the typical case, the topic of
the variance-quadrature rule of Volume I Chapter 4.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-error-bounds.tex}

\Cref{fig:vol02:ch01:error-bounds} draws the two accumulation curves for
$n$ equal contributions. The triangle inequality gives the straight
worst-case line $n\epsilon$, realised only when every error points the
same way. The quadrature curve $\sqrt{n}\,\epsilon$ sits far below it for
independent errors, because independent contributions partly cancel. The
gap is not a refinement of one bound by the other; the two answer
different questions. The engineer who must guarantee a specification
under adversarial alignment quotes the triangle bound; the engineer who
reports the expected spread of a measurement quotes the quadrature
estimate. Confusing the two either overstates a typical error by the
factor $\sqrt{n}$ or understates a worst case by the same factor.

\paragraph{Worked example: stacked tolerances on a shaft assembly.} Five
spacers are stacked on a shaft. Each spacer has nominal thickness
$10\,\si{\milli\meter}$ with an independent machining tolerance of $\pm
0.05\,\si{\milli\meter}$. The worst-case stack tolerance is the triangle
bound: all five tolerances at their extreme in the same direction gives
$5 \times 0.05 = 0.25\,\si{\milli\meter}$, so the stack height lies in
$50 \pm 0.25\,\si{\milli\meter}$. The statistical stack tolerance, valid
when the five deviations are independent and roughly centred, is the
quadrature estimate $\sqrt{5} \times 0.05 \approx
0.11\,\si{\milli\meter}$, so a typical stack lands in $50 \pm
0.11\,\si{\milli\meter}$. The two numbers differ by the factor
$\sqrt{5} \approx 2.24$. A designer sizing a mating clearance that must
never bind chooses the worst-case $0.25\,\si{\milli\meter}$; a designer
predicting the yield of a production run, where a rare all-aligned stack
is scrapped rather than catastrophic, works from the quadrature
$0.11\,\si{\milli\meter}$. The identity that licenses the smaller number
is the independence assumption, and the engineer who quotes the
quadrature stack without stating that the deviations are independent has
made a claim the triangle inequality does not support.

\subsection{The arithmetic mean--geometric mean inequality (AM-GM)}

For non-negative real numbers $a_{1}, \ldots, a_{n}$,
\[
\frac{a_{1} + a_{2} + \cdots + a_{n}}{n} \geq \sqrt[n]{a_{1} a_{2}
\cdots a_{n}},
\]
with equality if and only if $a_{1} = a_{2} = \cdots = a_{n}$.

The two-variable case is the most useful, and its proof is a single
algebraic step.

\begin{proof}[Proof of two-variable AM-GM]
For $a, b \geq 0$,
\[
(a - b)^{2} \geq 0
\]
expands to $a^{2} - 2ab + b^{2} \geq 0$. Add $4ab$ to both sides to
get $a^{2} + 2ab + b^{2} \geq 4ab$, that is, $(a+b)^{2} \geq 4ab$.
Take square roots (legal because both sides are non-negative) to
get $a + b \geq 2\sqrt{ab}$, hence $\frac{a+b}{2} \geq \sqrt{ab}$.
Equality holds when $(a-b)^{2} = 0$, that is, $a = b$.
\end{proof}

The general $n$-variable case follows by induction or by Jensen's
inequality applied to the concave function $\log$. We do not prove
the general case here. The two-variable case alone is the most
heavily used inequality in engineering optimisation: it tells the
engineer that the sum of two non-negative quantities, with their
product fixed, is minimised when the quantities are equal. Volume I
Chapter 7 used it implicitly to argue that a square minimises
perimeter for fixed area; later in this volume, Chapter 5 will use
it explicitly in the design of a cylindrical can with minimum
surface area for a fixed volume.

\begin{estimation}
\textbf{A closed cylindrical can must hold $V_{0} = 355
\,\si{\milli\liter}$, the volume of a standard beverage can.
Estimate, before computing, the radius $r$ that minimises the
total surface area $S = 2 \pi r^{2} + 2 \pi r h$, and the
fractional saving compared to the actual commercial design (typical
$r \approx 33 \,\si{\milli\meter}$, $h \approx 110 \,\si{\milli\meter}$).}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The constraint $\pi r^{2} h = V_{0}$ fixes $h$ once $r$ is chosen.
Substituting, the side area becomes $2 \pi r h = 2 V_{0} / r$, and
the total area is $S(r) = 2 \pi r^{2} + 2 V_{0} / r$. The two
contributions, top-plus-bottom (growing as $r^{2}$) and side (falling
as $1/r$), balance at the minimum. Setting $\dd S / \dd r = 4 \pi r
- 2 V_{0} / r^{2} = 0$ gives $r_{\min}^{3} = V_{0} / (2 \pi)$,
which the reader who prefers an AM--GM derivation can obtain by
splitting the side term into two equal halves and applying AM--GM
to the three positive terms $2 \pi r^{2}, V_{0}/r, V_{0}/r$; equality
in AM--GM asks $2 \pi r^{2} = V_{0}/r$, the same equation.

With $V_{0} = 355 \times 10^{-6} \,\si{\meter\cubed}$, $r_{\min}^{3}
= 5.65 \times 10^{-5} \,\si{\meter\cubed}$, so $r_{\min} \approx
3.83 \times 10^{-2} \,\si{\meter} \approx 38 \,\si{\milli\meter}$,
and the matching height is $h_{\min} = 2 r_{\min} \approx 77 \,\si
{\milli\meter}$. The closed-form minimum area is $S_{\min} = 3
\sqrt[3]{2 \pi V_{0}^{2}} \approx 2.77 \times 10^{-2}
\,\si{\meter\squared}$, against the actual $S_{\text{actual}} =
2 \pi r^{2} + 2 \pi r h \approx 2.96 \times 10^{-2} \,\si{\meter
\squared}$. The geometric saving is about $6 \%$. The beverage-can
designer reads this as: the surface-area optimum is shallow, and
the engineering reasons for departing from it (stacking, finger
ergonomics, pallet packing, internal-pressure-driven wall thickness,
branding area) cost only a few percent in metal. AM--GM tells the
designer where the optimum sits; the rest of the brief tells the
designer how far from the optimum to stand.
\end{estimation}

\subsection{Cauchy--Schwarz}

For real numbers $a_{1}, \ldots, a_{n}$ and $b_{1}, \ldots, b_{n}$,
\[
\left(\sum_{i=1}^{n} a_{i} b_{i}\right)^{2}
\leq \left(\sum_{i=1}^{n} a_{i}^{2}\right)
\left(\sum_{i=1}^{n} b_{i}^{2}\right),
\]
with equality if and only if the vectors $(a_{i})$ and $(b_{i})$ are
proportional. In vector form, $\abs{\inner{u, v}} \leq \norm{u}
\norm{v}$.

\begin{proof}
Consider the quadratic in $t$,
\[
Q(t) = \sum_{i=1}^{n} (a_{i} t - b_{i})^{2}
= \left(\sum a_{i}^{2}\right) t^{2}
- 2 \left(\sum a_{i} b_{i}\right) t
+ \sum b_{i}^{2}.
\]
$Q(t) \geq 0$ for every real $t$, so the discriminant of $Q$ is
non-positive,
\[
4 \left(\sum a_{i} b_{i}\right)^{2}
- 4 \left(\sum a_{i}^{2}\right) \left(\sum b_{i}^{2}\right) \leq 0,
\]
which rearranges to the stated inequality. Equality holds when
$Q(t)$ has a real double root, that is, when $a_{i} t = b_{i}$ for
all $i$ and some real $t$; this is the proportionality condition.
\end{proof}

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-cauchy-schwarz-geom.tex}

\Cref{fig:vol02:ch01:cauchy-schwarz} draws the inequality in the
plane. The inner product $\inner{u, v}$ is the signed length of the
projection of $v$ onto $u$, scaled by $\norm{u}$; that projected
length is bounded above by $\norm{v}$ itself, which is the geometric
reading of the bound.

Cauchy--Schwarz is the inequality that makes inner products geometric.
The proof we have given is the textbook one; an alternative proof
through the AM--GM inequality applied term by term is in the
exercises. The result is the bound on a correlation coefficient
(Volume I Chapter 8 stated $-1 \leq \rho \leq 1$ without proof; that
bound is Cauchy--Schwarz applied to centred random variables), the
bound on power dissipation in a circuit, the foundation of the
projection formula in Chapter 8 of this volume, and the variational
bound that supports much of optimisation in Chapter 15.

\paragraph{Applied vignette: the power factor as a Cauchy--Schwarz
ratio.} An alternating-current load draws instantaneous current $i(t)$
under an applied voltage $v(t)$ over one period $[0, T]$. Three
quantities describe the load. The real power delivered is the average of
the instantaneous product, $P = \frac{1}{T}\int_{0}^{T} v(t)\, i(t)\,
\dd t$. The apparent power is the product of the root-mean-square voltage
and current, $S = V_{\mathrm{rms}} I_{\mathrm{rms}}$, where
$V_{\mathrm{rms}}^{2} = \frac{1}{T}\int_{0}^{T} v^{2}\,\dd t$ and
similarly for the current. The power factor is the ratio $\mathrm{pf} =
P/S$, and every utility engineer knows it cannot exceed one. The bound
is Cauchy--Schwarz for functions. Treating the time integral as the
inner product $\inner{v, i} = \frac{1}{T}\int_{0}^{T} v i\, \dd t$, the
inequality reads
\[
\left(\frac{1}{T}\int_{0}^{T} v\, i\, \dd t\right)^{2}
\leq
\left(\frac{1}{T}\int_{0}^{T} v^{2}\, \dd t\right)
\left(\frac{1}{T}\int_{0}^{T} i^{2}\, \dd t\right),
\]
that is, $P^{2} \leq V_{\mathrm{rms}}^{2} I_{\mathrm{rms}}^{2} = S^{2}$,
so $\abs{\mathrm{pf}} \leq 1$. The equality condition is the content the
engineer actually uses. Cauchy--Schwarz is tight exactly when the two
functions are proportional, $i(t) = \kappa\, v(t)$ for a constant
$\kappa$, which is the case of a purely resistive load: current in phase
with voltage, power factor one, every volt-ampere delivered as real
power. Any phase lag or harmonic distortion makes $i$ non-proportional to
$v$, drops the power factor below one, and the shortfall $S^{2} - P^{2}$
is the reactive-plus-distortion power the utility must supply but cannot
bill as work. For the textbook sinusoidal case $v = V_{0}\cos\omega t$,
$i = I_{0}\cos(\omega t - \phi)$, the integral gives $P = \tfrac12 V_{0}
I_{0}\cos\phi$ and $S = \tfrac12 V_{0} I_{0}$, so $\mathrm{pf} =
\cos\phi$, and the Cauchy--Schwarz bound $\cos\phi \leq 1$ is the
familiar statement that the phase angle can only cost, never add. The
same inequality that bounds a correlation coefficient by one, and that
bounds the length of a projection by the length of the vector, bounds
the fraction of apparent power a load turns into work. One theorem, three
readings.

\subsection{The power-mean inequality}

AM-GM is one rung of a longer ladder. For positive reals $a_{1},
\ldots, a_{n}$ and a real exponent $p \neq 0$, the power mean of order
$p$ is
\[
M_{p} = \left(\frac{1}{n} \sum_{i=1}^{n} a_{i}^{p}\right)^{1/p},
\]
with $M_{0}$ defined by the limit $p \to 0$, which equals the
geometric mean. The power mean is non-decreasing in $p$: for $p < q$,
$M_{p} \leq M_{q}$, with equality if and only if all the $a_{i}$ are
equal. The familiar means are particular orders. The harmonic mean is
$M_{-1}$, the geometric mean is $M_{0}$, the arithmetic mean is $M_{1}$,
and the quadratic mean (root mean square) is $M_{2}$. The chain
\[
\mathrm{HM} \leq \mathrm{GM} \leq \mathrm{AM} \leq \mathrm{QM}
\]
is the power-mean inequality read at four specific orders, and AM-GM
is the single step from $M_{0}$ to $M_{1}$.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-power-mean.tex}

\Cref{fig:vol02:ch01:power-mean} plots the power mean of the two
numbers $2$ and $8$ as a function of $p$, with the four classical
means marked. The curve runs monotonically from $\min(a_{i})$ as
$p \to -\infty$ to $\max(a_{i})$ as $p \to +\infty$, passing through
the harmonic, geometric, arithmetic, and quadratic means in order. The
engineering uses are concrete. The harmonic mean is the correct
average for rates combined in series, such as the effective resistance
of parallel resistors or the average speed over a fixed distance at
two speeds. The quadratic mean is the correct average for quantities
whose energy or power scales as the square, such as the root-mean-
square value of an alternating current that the reader meets in
Chapter 3. The arithmetic mean is the default that an engineer reaches
for without thinking. The power-mean inequality is the formal warning
that these are different numbers, ordered, and that reaching for the
arithmetic mean when the harmonic mean is required overstates the
answer by a known amount.

\paragraph{Worked example: average speed is the harmonic mean.}
A delivery van covers the same route segment at $30 \,\si{\kilo\meter
\per\hour}$ outbound and $60 \,\si{\kilo\meter\per\hour}$ return. The
average speed over the round trip is not the arithmetic mean $45
\,\si{\kilo\meter\per\hour}$. Equal distance at two speeds averages by
the harmonic mean,
\[
\bar v = \frac{2}{\tfrac{1}{30} + \tfrac{1}{60}}
= \frac{2}{\tfrac{3}{60}} = 40 \,\si{\kilo\meter\per\hour}.
\]
The harmonic mean ($M_{-1} = 40$) sits below the arithmetic mean
($M_{1} = 45$), exactly as the power-mean inequality requires. The
engineer who quotes $45$ has averaged the wrong quantity; the trip
spends more time at the slower speed, and time is what the average must
weight.

\subsection{Young's inequality and weighted AM-GM}

A weighted version of AM-GM underwrites much of optimisation. For
positive reals $a, b$ and weights $\lambda, \mu \geq 0$ with $\lambda
+ \mu = 1$,
\[
\lambda a + \mu b \geq a^{\lambda} b^{\mu},
\]
with equality if and only if $a = b$. The unweighted AM-GM is the case
$\lambda = \mu = \tfrac{1}{2}$. Setting $\lambda = 1/p$ and $\mu = 1/q$
with $\tfrac{1}{p} + \tfrac{1}{q} = 1$ and renaming gives Young's
inequality,
\[
xy \leq \frac{x^{p}}{p} + \frac{y^{q}}{q},
\qquad x, y \geq 0, \quad p, q > 1, \quad \frac{1}{p} + \frac{1}{q} = 1,
\]
which is the inequality that generates Hölder's inequality and, with
$p = q = 2$, reduces to the AM-GM bound $xy \leq \tfrac{1}{2}(x^{2} +
y^{2})$ used in the Cauchy-Schwarz derivation of the exercises. Young's
inequality is the tool an engineer reaches for to split a product into
a controllable sum of powers, which is the recurring move in bounding
error terms and in the convex-duality arguments of Chapter 15.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-young-inequality.tex}

\Cref{fig:vol02:ch01:young-inequality} reads Young's inequality as an
area balance, which is the picture that makes the exponents $p$ and $q$
concrete. The product $ab$ is the area of a rectangle, and it is covered
by the area under a monotone curve plus the area to the left of it, with
equality only when the corner of the rectangle sits on the curve. The
picture also states the equality condition without algebra: the bound is
tight exactly when $b = f(a)$, which for $f(x) = x^{p-1}$ is the
condition $a^{p} = b^{q}$, the weighted-AM--GM equality in disguise. An
engineer who splits a troublesome product $ab$ into $a^{p}/p + b^{q}/q$
chooses the exponent $p$ to make one of the two terms small enough to
absorb, and the picture shows what that choice costs: moving $p$ away
from the value that balances the two areas widens the gap between the
rectangle and the curve.

\begin{mastery}
The monotonicity of the power mean is one instance of Jensen's
inequality, which states that for a convex function $\varphi$ and
weights $w_{i} \geq 0$ summing to one,
\[
\varphi\!\left(\sum_{i} w_{i} a_{i}\right)
\leq \sum_{i} w_{i} \varphi(a_{i}).
\]
Applying Jensen to the convex function $\varphi(t) = t^{q/p}$ (for
$q > p > 0$) on the values $a_{i}^{p}$ yields $M_{p} \leq M_{q}$;
applying it to $\varphi = -\log$ yields AM-GM directly. Jensen's
inequality is the single statement from which the entire family of
mean inequalities descends, and its proof rests on the definition of
convexity, which the reader meets formally in Chapter 5 and uses
heavily in the optimisation chapter. The engineering reading of Jensen
is that averaging commutes with a convex function only as an
inequality, never as an equality unless the inputs are constant. An
engineer who computes the average of a nonlinear function of a noisy
input, and reports it as the function of the average input, has made a
Jensen error: the two differ by an amount that grows with the input
variance and the curvature of the function. This is the same
distinction that separates the mean of a ratio from the ratio of
means, a trap the reader meets in error propagation in Volume I
Chapter 4.
\end{mastery}

\subsection{The reverse triangle inequality}

A small consequence of the triangle inequality is heavily used in
error analysis,
\[
\abs{\,\abs{a} - \abs{b}\,} \leq \abs{a - b}.
\]
The reverse triangle inequality says that the difference of two
magnitudes is at most the magnitude of the difference. The
engineering reading is that small absolute errors in two readings
produce at most a small absolute error in the difference of their
magnitudes; the size of the readings themselves does not amplify
the error. The result is the bound the engineer wants when checking
that a calibration drift has not corrupted a magnitude measurement.

\subsection{Worked bound: relative error of a product and a quotient}

The inequalities of this section earn their place when they convert a
vague worry about accumulated error into a stated bound. We work one
case in full, because the result is used without comment in Volume I
Chapter 4 and deserves a derivation the reader can audit. The claim is
that the relative error of a product is, to first order, bounded by the
sum of the relative errors of the factors, and that the same holds for a
quotient.

\begin{theorem}
Let $\tilde a = a(1 + \alpha)$ and $\tilde b = b(1 + \beta)$ be
measured values of nonzero quantities $a$ and $b$, with relative errors
$\abs{\alpha} \leq \epsilon_{a}$ and $\abs{\beta} \leq \epsilon_{b}$.
Then the relative error of the product satisfies
\[
\left\lvert \frac{\tilde a \tilde b - ab}{ab} \right\rvert
\leq \epsilon_{a} + \epsilon_{b} + \epsilon_{a}\epsilon_{b},
\]
and the relative error of the quotient, when $\epsilon_{b} < 1$,
satisfies
\[
\left\lvert \frac{\tilde a / \tilde b - a/b}{a/b} \right\rvert
\leq \frac{\epsilon_{a} + \epsilon_{b}}{1 - \epsilon_{b}}.
\]
\end{theorem}

\begin{proof}
For the product, expand directly:
\[
\frac{\tilde a \tilde b}{ab}
= (1 + \alpha)(1 + \beta)
= 1 + \alpha + \beta + \alpha\beta,
\]
so the relative error is $\alpha + \beta + \alpha\beta$. By the triangle
inequality, $\abs{\alpha + \beta + \alpha\beta} \leq \abs{\alpha} +
\abs{\beta} + \abs{\alpha}\abs{\beta} \leq \epsilon_{a} + \epsilon_{b} +
\epsilon_{a}\epsilon_{b}$, which is the first bound. The cross term
$\epsilon_{a}\epsilon_{b}$ is second order: when both relative errors
are one percent it contributes one part in ten thousand, which is why
the first-order reading drops it and reports the relative error of a
product as the sum of the relative errors of the factors.

For the quotient,
\[
\frac{\tilde a / \tilde b}{a/b}
= \frac{1 + \alpha}{1 + \beta}.
\]
Write the relative error as
\[
\frac{1 + \alpha}{1 + \beta} - 1
= \frac{(1 + \alpha) - (1 + \beta)}{1 + \beta}
= \frac{\alpha - \beta}{1 + \beta}.
\]
The numerator is bounded by $\abs{\alpha} + \abs{\beta} \leq
\epsilon_{a} + \epsilon_{b}$ by the triangle inequality. The
denominator satisfies $\abs{1 + \beta} \geq 1 - \abs{\beta} \geq 1 -
\epsilon_{b} > 0$ by the reverse triangle inequality and the hypothesis
$\epsilon_{b} < 1$. Dividing the bound on the numerator by the lower
bound on the denominator preserves the inequality because both are
positive, giving the stated quotient bound.
\end{proof}

The proof uses three of this section's tools in sequence: the triangle
inequality to bound the numerator, the reverse triangle inequality to
lower-bound the denominator, and the sign rule for dividing an
inequality by a positive quantity. The hypothesis $\epsilon_{b} < 1$ is
the licensing condition the cancellation discussion of section 1.1
insisted on; without it the denominator can change sign and the bound is
false. The engineering reading is the rule of thumb every working
engineer carries: relative errors add under multiplication and division.
The derivation shows the rule is a first-order truncation of an exact
inequality, that the dropped term is second order, and that the quotient
form needs the divisor's error to stay below unity. Volume I Chapter 4
builds the variance-quadrature refinement on top of this worst-case
bound, replacing the sum of relative errors by the square root of the
sum of squared relative errors when the errors are independent; the two
bounds answer different questions, the worst case and the typical case,
and the engineer chooses between them by whether the errors can conspire.

\subsection{Bernoulli's inequality}

For $x \geq -1$ and $n$ a non-negative integer,
\[
(1 + x)^{n} \geq 1 + nx.
\]
The proof is by induction (section 1.4). The inequality is loose for
moderate $x$ and sharp for small $x$, where the binomial theorem's
$1 + nx + \tfrac{n(n-1)}{2} x^{2} + \cdots$ shows that the error in
the linear approximation grows like $x^{2}$. Engineers use Bernoulli
as the conservative version of the linear approximation: it gives a
guaranteed lower bound where the linear approximation gives only a
typical value. The conservative bound is what an engineer wants in a
worst-case analysis.

\subsection{When inequalities go the wrong way}

The most common error in inequality manipulation is reversing the
direction by accident. Three rules guard against this. Multiplying
or dividing both sides by a positive number preserves the inequality.
Multiplying or dividing by a negative number reverses it. Squaring
preserves the inequality only when both sides are non-negative; for
arbitrary real numbers, $a < b$ does not imply $a^{2} < b^{2}$
(consider $a = -2$, $b = 1$). Taking reciprocals reverses the
inequality on the same-sign part of the real line, and is undefined
across zero.

A reader who manipulates inequalities should re-state the sign and
domain of the multiplier or function at every step. The discipline
is the same as for cancellation: stating the licensing condition is
what makes the manipulation defensible.

\section{What a proof is, what it costs, why engineers occasionally
need them}\pathtag{standard}

A proof is an argument that establishes a claim from explicitly
stated premises by steps each of which a reader can verify. A
calculation can be wrong because of an arithmetic slip; a proof can
be wrong because of a logical slip; the difference is that a proof's
slip cannot be repaired by checking the arithmetic. The engineer who
wants to know whether a claim is true under all conditions stated,
not just under the conditions tested, eventually needs a proof.

\subsection{What a proof costs}

A proof is expensive. It requires the writer to fix the premises in
advance, to articulate every step, to refuse the appeal to intuition
where intuition is what the reader is being asked to set aside, and
to write the argument in a form that another reader can audit. The
expense is justified when the claim is load-bearing across a domain
in which experiment is too costly, too dangerous, or too partial to
substitute. A safety property of a piece of cryptographic code, a
stability property of a numerical scheme, a uniqueness property of
a solution to an ODE, an upper bound on the worst-case running time
of an algorithm; in each of these the engineer cannot afford to know
the property only experimentally. The engineer needs to know the
property holds.

A proof is also expensive because it forces the writer to confront
the gap between what they think they know and what they can write
down. A common experience in proof writing is the discovery that a
step the writer thought was obvious is in fact false, or that a
hypothesis the writer thought was harmless is in fact necessary. The
discipline of writing the proof is what surfaces these.

\subsection{What a proof is not}

A proof is not a derivation. A derivation manipulates symbols toward
a conclusion; a proof argues that the conclusion follows from the
premises. The two often look alike on the page; the difference is
the writer's intent and the standard the writer holds the argument
to.

A proof is not a demonstration in the experimental sense. Running an
algorithm on ten thousand inputs and seeing it produce the right
answer is a demonstration. It is not a proof. The eleven-thousandth
input may be the input on which the algorithm fails, and the
demonstration cannot rule that out.

A proof is not a heuristic. A heuristic is a productive guess about
where to look. Heuristics are essential to mathematical work, as
P\'{o}lya's \emph{How to Solve It} \cite{gen:polya1945} argues at
book length, but a heuristic does not establish a result.

\begin{estimation}
\textbf{A safety-critical function takes a single 32-bit integer
input. Exhaustive testing would run the function on every possible
input and check the output. At a sustained rate of $10^{9}$ test cases
per second, estimate, before computing, how long exhaustive testing
takes for a 32-bit input, and then for a 64-bit input.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A 32-bit input has $2^{32} \approx 4.3 \times 10^{9}$ possible values.
At $10^{9}$ cases per second the run takes about $4.3$ seconds, which
is feasible: a 32-bit input can in fact be tested exhaustively, and
for such a function a demonstration over all inputs is a proof, because
the input space is finite and small. A 64-bit input has $2^{64}
\approx 1.8 \times 10^{19}$ values. At the same rate the run takes
$1.8 \times 10^{19} / 10^{9} = 1.8 \times 10^{10}$ seconds, which is
about $580$ years. Exhaustive testing of a 64-bit input is infeasible,
and a demonstration over any reachable sample covers a vanishing
fraction of the space. The doubling of the input width from 32 to 64
bits moves the function from the regime where testing can be exhaustive
to the regime where only a proof can establish a property over all
inputs. The estimate is the quantitative form of the chapter's
distinction: the eleven-thousandth input matters precisely when the
input space is too large to enumerate, and a single integer's width
decides which regime the engineer is in.
\end{estimation}

\subsection{The standard proof techniques}

Five techniques cover most engineering proofs.

\engterm{Direct proof}. Assume the hypothesis; derive the conclusion
through a chain of valid steps. The triangle inequality proof of
section 1.2 was direct.

\engterm{Proof by cases}. Partition the hypothesis into cases that
exhaust the possibilities; prove the conclusion in each case. The
triangle inequality proof used cases on the sign of $a + b$.

\engterm{Proof by contradiction}. Assume the hypothesis and the
negation of the conclusion; derive a contradiction; conclude that the
hypothesis must imply the conclusion. The classical proof that
$\sqrt{2}$ is irrational is by contradiction.

\engterm{Proof by contrapositive}. To prove ``if $P$ then $Q$,''
prove ``if not $Q$ then not $P$.'' The contrapositive is logically
equivalent to the original, and is often easier to handle when the
direct version requires constructing an existence witness.

\engterm{Proof by induction}. To prove a statement for every natural
number, prove the base case and the inductive step. Section 1.4
treats induction in detail.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-proof-techniques.tex}

\Cref{fig:vol02:ch01:proof-techniques} arranges the five techniques as
a decision tree the writer can walk before reaching for algebra. The
tree gives a starting move, not a guarantee; most real proofs combine
techniques, and the value of the tree is that it forces the writer to
name the logical shape of the claim, universally quantified or
existential, over the natural numbers or over a continuum, before
writing the first line.

\subsection{Worked proof by contrapositive}

The contrapositive turns a claim that is awkward to prove directly
into one that is direct. We take a small number-theoretic claim as the
example because the direct version would require constructing a factor
and the contrapositive does not.

\begin{theorem}
For an integer $n$, if $n^{2}$ is even then $n$ is even.
\end{theorem}

\begin{proof}
The contrapositive of ``if $n^{2}$ is even then $n$ is even'' is ``if
$n$ is odd then $n^{2}$ is odd,'' which is logically equivalent. Assume
$n$ is odd. Then $n = 2k + 1$ for some integer $k$, and $n^{2} = 4k^{2}
+ 4k + 1 = 2(2k^{2} + 2k) + 1$, which is odd. The contrapositive holds,
hence the original.
\end{proof}

The direct proof would assume $n^{2}$ even and try to exhibit $n$ as
twice an integer, which requires extracting a factor of two from a
square root and is more delicate than the one-line contrapositive.
This small lemma is the step the $\sqrt{2}$ proof below relies on, and
the contrapositive is how that step is most cleanly established.

\subsection{Worked proof by cases}

Proof by cases partitions the hypothesis into a finite set of
possibilities that exhaust it and proves the conclusion in each. The
discipline is to check that the cases genuinely cover everything, because
a missed case leaves a hole no amount of care inside the other cases can
fill. We prove a small fact an engineer uses without thinking whenever
rounding to the nearest integer.

\begin{theorem}
For every real number $x$, the nearest integer to $x$ differs from $x$
by at most $\tfrac12$; that is, there is an integer $m$ with $\abs{x - m}
\leq \tfrac12$.
\end{theorem}

\begin{proof}
Write $x = n + f$ where $n = \lfloor x \rfloor$ is the integer part and
$f = x - n$ is the fractional part, so $0 \leq f < 1$. We argue by cases
on $f$. Case one, $f \leq \tfrac12$: take $m = n$. Then $\abs{x - m} = f
\leq \tfrac12$, as required. Case two, $f > \tfrac12$: take $m = n + 1$.
Then $\abs{x - m} = \abs{(n + f) - (n + 1)} = 1 - f$, and $f > \tfrac12$
gives $1 - f < \tfrac12$, so $\abs{x - m} < \tfrac12$. The two cases
cover every $f$ in $[0, 1)$ because every real $f$ is either at most
$\tfrac12$ or greater than $\tfrac12$, so the conclusion holds for every
$x$.
\end{proof}

The proof is complete because the two cases partition the range of $f$
with no gap and no overlap at the decision point (the boundary $f =
\tfrac12$ is assigned to case one). The engineering reading is the
quantisation bound of an ideal rounder: the worst-case rounding error of
round-to-nearest is exactly half a least significant unit, and the case
split is where that half comes from. A rounder that broke the cases
differently, say by rounding $f = \tfrac12$ down every time rather than
to even, would still meet the same $\tfrac12$ bound; the case analysis
shows the bound is a property of the partition, not of the tie-breaking
rule.

\subsection{Worked proof of existence and uniqueness}

Engineers care about existence and uniqueness because a model with no
solution is unphysical and a model with many solutions is
underdetermined. The simplest worked case is the solution of a linear
equation, which previews the existence-and-uniqueness questions that
dominate Chapter 9.

\begin{theorem}
For real numbers $a \neq 0$ and $b$, the equation $a x = b$ has exactly
one real solution.
\end{theorem}

\begin{proof}
Existence: $x = b/a$ is defined because $a \neq 0$, and substituting
gives $a (b/a) = b$, so a solution exists. Uniqueness: suppose
$x_{1}$ and $x_{2}$ both solve the equation. Then $a x_{1} = b =
a x_{2}$, so $a (x_{1} - x_{2}) = 0$. Because $a \neq 0$ and the reals
have no zero divisors, $x_{1} - x_{2} = 0$, that is, $x_{1} = x_{2}$.
The solution is unique.
\end{proof}

The proof has the canonical two-part shape: existence is a
construction, uniqueness is a contradiction of having two distinct
solutions. The hypothesis $a \neq 0$ does double duty, licensing the
division in existence and the cancellation in uniqueness. When $a = 0$
the theorem is false in both directions: $0 \cdot x = b$ has no
solution if $b \neq 0$ and infinitely many if $b = 0$. The reader will
meet exactly this trichotomy, with $a$ replaced by a matrix and the
zero-divisor argument replaced by invertibility, in the linear-algebra
chapter. The shape of the proof carries across the change of objects.

\subsection{Worked proof: $\sqrt{2}$ is irrational}

We take the classical proof as a worked example because it shows
how cleanly proof by contradiction can establish a non-existence
claim that no amount of computation could establish.

\begin{theorem}
$\sqrt{2}$ is irrational.
\end{theorem}

\begin{proof}
Suppose for contradiction that $\sqrt{2}$ is rational. Then we can
write $\sqrt{2} = p/q$ for integers $p$ and $q$ with $q \neq 0$ and
$\gcd(p, q) = 1$ (any rational has a representation in lowest
terms). Squaring, $2 = p^{2}/q^{2}$, that is, $p^{2} = 2 q^{2}$.

Therefore $p^{2}$ is even. The square of an odd integer is odd, so
$p$ itself is even. Write $p = 2 m$ for some integer $m$. Then
$p^{2} = 4 m^{2}$, and substituting into $p^{2} = 2 q^{2}$ gives
$4 m^{2} = 2 q^{2}$, that is, $q^{2} = 2 m^{2}$. By the same
argument, $q$ is even.

Both $p$ and $q$ are even. But this contradicts the assumption
$\gcd(p, q) = 1$. The assumption that $\sqrt{2}$ is rational is
therefore false, and $\sqrt{2}$ is irrational.
\end{proof}

The proof has three features the reader should notice. The
hypothesis $\gcd(p, q) = 1$ is not optional; without it the
contradiction does not arise. The step ``the square of an odd
integer is odd'' is itself a small lemma, which the reader can
verify by writing $(2k+1)^{2} = 4k^{2} + 4k + 1$. The proof
establishes a result that no finite computation could: that no
integer fraction equals $\sqrt{2}$. Decimal computation can show
$\sqrt{2}$ is approximately $1.414213\ldots$ to any number of
digits, but no finite decimal expansion is the same as proving no
fraction matches.

\subsection{Worked proof: $\log_{2} 3$ is irrational}

A second contradiction proof lands on a claim an engineer meets in
information theory and in the analysis of algorithms, where logarithms
base two count bits and comparisons. The proof uses the unique
factorisation of integers rather than a parity argument, which shows the
contradiction move working through a different property of the integers.

\begin{theorem}
$\log_{2} 3$ is irrational.
\end{theorem}

\begin{proof}
Suppose for contradiction that $\log_{2} 3$ is rational, $\log_{2} 3 =
p/q$ for positive integers $p$ and $q$. By the definition of the
logarithm, $2^{p/q} = 3$, and raising both sides to the $q$ gives
$2^{p} = 3^{q}$. The left side is a power of two, so it is even for every
$p \geq 1$; the right side is a power of three, so it is odd for every
$q \geq 0$. An integer cannot be both even and odd, so $2^{p} = 3^{q}$ is
impossible for positive $p$ and $q$. The assumption that $\log_{2} 3$ is
rational is therefore false.
\end{proof}

The sharper reading uses unique factorisation directly: $2^{p}$ has only
the prime $2$ in its factorisation and $3^{q}$ has only the prime $3$, so
the two can be equal only when both exponents are zero, which the
positivity of $p$ and $q$ forbids. The engineering payoff is that a
quantity like $\log_{2} 3 \approx 1.585$ bits, the information content of
one ternary symbol measured in bits, is not any ratio of integers; a
coding scheme that packs ternary symbols into a whole number of bits per
symbol can only approach the rate $\log_{2} 3$, never meet it exactly,
and the shortfall is the reason block coding over many symbols is
required to approach the entropy limit of Volume VII. The irrationality
is not a curiosity; it is the formal reason a per-symbol integer bit
budget always leaves rate on the table.

\subsection{Why engineers occasionally need them}

Most engineering work is calculation, derivation, and measurement.
Most of the time, an engineer who can do those three carefully has
done what is required. Proofs enter when the claim is one whose
violation would not show up in the calculations the engineer ran.

A short list of cases. The engineer designing a numerical scheme
needs to know it is stable for every initial condition in a stated
class, not just the ones that happen to have been tested. The
engineer specifying a cryptographic protocol needs to know the
protocol resists every attack from a stated class of attackers, not
just the attacks that have been imagined. The engineer arguing the
correctness of a fault-tolerant algorithm needs to know the algorithm
remains correct under every interleaving of failures, not just the
interleavings that happen to have been simulated. In each case, the
proof is not a luxury; it is the only way the engineer can defend
the property in front of a sceptical audience.

The reverse case is also worth stating. Many engineering claims are
empirical, contingent, dated, or domain-specific. A proof is the
wrong tool for ``this material works at this temperature,'' for ``this
process produces parts within tolerance,'' for ``this road carries
this volume of traffic.'' Demanding a proof of such a claim is a
category error. The engineer's working judgement includes knowing
which claims are amenable to proof and which are not.

\subsection{Worked proof: there are infinitely many primes}

A second non-existence proof shows the same contradiction move on a
different claim, and is worth seeing because the construction inside it
is the model for many later existence arguments.

\begin{theorem}
There are infinitely many prime numbers.
\end{theorem}

\begin{proof}
Suppose for contradiction that there are finitely many primes, say
$p_{1}, p_{2}, \ldots, p_{m}$, the complete list. Form the number
$N = p_{1} p_{2} \cdots p_{m} + 1$. $N$ is larger than every prime on
the list, so it is not itself on the list, and by the supposed
completeness of the list $N$ is not prime. Therefore $N$ has a prime
divisor $p$, and $p$ must be one of $p_{1}, \ldots, p_{m}$. But $p$
divides the product $p_{1} \cdots p_{m}$, and $p$ divides $N$, so $p$
divides their difference $N - p_{1} \cdots p_{m} = 1$. No prime
divides $1$. The contradiction shows the supposition false, so the
primes are infinite.
\end{proof}

The construction of $N$ is the heart of the proof: from any finite
collection it manufactures an object that the collection cannot
contain. This manufacture-a-witness move recurs in many existence
proofs, including the diagonal argument the mastery box below sketches
and the pigeonhole arguments of Volume VII. The proof does not produce
the next prime; it only proves one exists beyond any finite list. The
distinction between a proof of existence and a construction of the
object is one an engineer should keep sharp, because a non-constructive
existence proof guarantees a solution without telling the engineer how
to compute it.

\subsection{The counterexample: the cheapest disproof}

A proof establishes a universal claim; a single counterexample destroys
one. The asymmetry is the engineer's most economical tool, because
finding one input on which a claimed property fails costs far less than
proving the property holds on all of them, and the engineer who suspects
a claim is false should look for the counterexample before attempting
the proof. The discipline is to read a universally quantified claim,
``for every $x$, $P(x)$,'' as an invitation to search for the one $x$
where $P$ breaks. A claim that survives a serious search for a
counterexample is worth the cost of a proof; a claim that falls to the
first odd case never needed one.

Three worked counterexamples show the move. The claim ``$n^{2} + n + 41$
is prime for every non-negative integer $n$'' holds for $n = 0$ through
$n = 39$, which is long enough to tempt belief, and fails at $n = 40$,
where $40^{2} + 40 + 41 = 1681 = 41^{2}$. Forty confirmations are not a
proof, and the forty-first input is the counterexample. The claim ``if
$a < b$ then $a^{2} < b^{2}$'' fails at $a = -2$, $b = 1$, the
counterexample the inequality section already used to warn against
squaring across a sign change. The claim ``a function continuous on an
open interval attains a maximum on it'' fails for $f(x) = x$ on $(0,
1)$, which has a supremum it never reaches; the counterexample is why the
maximum theorems the reader meets later insist on a closed interval. In
each case the counterexample does in one line what no number of
confirming cases could do, and the engineering reading is that the
demonstration over a sample, however large, is refuted by a single
adverse input. The relationship to testing is exact: a test that fails
is a counterexample to the claim that the software is correct, and is
conclusive, while a test that passes confirms nothing universal. The
engineer who internalises the asymmetry tests to disprove, designs the
adversarial input first, and reserves the expense of proof for the
claims that have earned it by surviving the search.

\begin{mastery}
The \engterm{pigeonhole principle} is the shortest existence proof in the
engineer's kit: if $n + 1$ objects are placed in $n$ boxes, some box
holds at least two objects. The statement is trivial and the
consequences are not, because it manufactures an existence guarantee from
a counting mismatch alone, with no construction of the object it
promises. The proof is one line by contradiction: if every box held at
most one object, the total would be at most $n$, contradicting $n + 1$.
The engineering uses are exact counting arguments. A hash table that maps
more than $n$ distinct keys into $n$ buckets must produce a collision;
the pigeonhole principle proves collisions are unavoidable once the key
count exceeds the bucket count, before any particular hash function is
examined, which is why a collision-handling strategy is not optional. A
lossless compressor cannot shorten every input: the strings of length
$n$ number $2^{n}$, the strings shorter than $n$ number $2^{n} - 1$, so
no injective map sends every length-$n$ string to a shorter one, and some
input must grow, the counting proof that a universal compressor is
impossible. An analog-to-digital converter with $2^{N}$ output codes,
fed more than $2^{N}$ distinguishable input levels, must assign two
levels the same code, which is the quantisation collision of Volume~I
Chapter~6 stated as a pigeonhole fact. The generalised form, that $m$
objects in $n$ boxes force some box to hold at least $\lceil m/n \rceil$,
sharpens the bound. The principle recurs in the complexity and coding
arguments of Volume~VII, and its value to an engineer is that it settles
an existence question by arithmetic, with no need to exhibit the colliding
pair.
\end{mastery}

\subsection{Proof, software, and formal verification}\pathtag{mastery}

The engineering setting in which proof has become routine is the
verification of safety-critical and security-critical software. A
proof that a numerical scheme is stable, that a concurrent algorithm
is free of deadlock, or that a cryptographic protocol resists a class
of attacks is increasingly written in a form a machine can check.
Tools such as model checkers, SMT solvers, and proof assistants take
the burden of auditing each step off the human reader and place it on
software whose own correctness has itself been argued. The cost
structure of a proof shifts under these tools: the up-front cost of
formalising the claim rises, and the per-revision cost of re-checking
the proof after a change falls to nearly zero. For a component that
will be revised many times over a long deployment, the formal proof
amortises in a way the hand proof does not.

The limit of the approach is the gap between the formal model and the
physical system. A machine-checked proof establishes a property of the
model, not of the silicon, the network, or the operator. The Ariane 5
software of section 1.7 was correct against its Ariane 4 model; the
model's flight envelope was the unstated assumption that the proof, had
one existed, would have inherited. Formal verification raises the floor
on what an engineer can claim with confidence and moves the residual
risk to the boundary between model and world. That boundary is where
the chapter's interface discipline lives, and it is the boundary no
proof can cross on its own.

\begin{mastery}
Cantor's diagonal argument is the canonical example of a proof that
establishes a result no computation could reach, and it is worth one
paragraph because its shape recurs in undecidability and complexity
results the reader meets in Volume VII. The claim is that the real
numbers in $[0, 1]$ cannot be put in a one-to-one correspondence with
the natural numbers; the set of reals is, in the precise sense,
strictly larger. Suppose a list pairs each natural number $n$ with a
real $r_{n}$, written in decimal. Construct a new real $d$ whose
$n$-th decimal digit differs from the $n$-th digit of $r_{n}$ (for
instance, by adding one mod ten, avoiding $9$ and $0$ to dodge the
$0.999\ldots$ ambiguity). Then $d$ differs from every $r_{n}$ in at
least one digit, so $d$ is not on the list, contradicting the
supposition that the list was complete. The argument manufactures a
witness from the supposed-complete list, exactly as the
infinitude-of-primes proof did. The same diagonalisation, applied to
programs rather than reals, proves the halting problem undecidable. An
engineer does not need the set theory, but the working lesson carries:
some collections are too large to enumerate, and a proof can establish
that fact about an infinite object that no finite test could approach.
\end{mastery}

\section{Induction}\pathtag{standard}

Induction is the proof technique that establishes a statement
$P(n)$ for every natural number $n$. The technique has two parts.
The base case proves $P(1)$ (or $P(0)$, depending on the
formulation). The inductive step proves that for every $n$, if
$P(n)$ holds then $P(n+1)$ holds. Together the two parts establish
$P(n)$ for every $n \geq 1$.

\subsection{Why induction works}

The principle of induction is one of the axioms of the natural
numbers; it does not derive from the others. The intuition is the
falling-domino picture: if the first domino falls, and if every
falling domino knocks down its successor, then every domino falls.
The mathematical content is the same. Engineers occasionally meet
this principle uncomfortably because it lets a finite argument cover
infinitely many cases, and an engineer's intuition is built on
finite testing.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-induction-dominoes.tex}

\Cref{fig:vol02:ch01:induction-dominoes} draws the two halves of the
argument and shows why neither alone suffices. A base case with no
inductive step tips one tile and leaves the rest standing. An inductive
step with no base case is a row of tiles each ready to tip its
successor, with nothing to tip the first. Both halves are load-bearing,
and the most common inductive error, treated below, is to lose one of
them without noticing.

\subsection{Worked example: sum of the first $n$ integers}

\begin{theorem}
For every natural number $n \geq 1$,
\[
\sum_{i=1}^{n} i = \frac{n(n+1)}{2}.
\]
\end{theorem}

\begin{proof}
Base case: $n = 1$. The left side is $1$. The right side is $\tfrac{1
\cdot 2}{2} = 1$. Both sides equal $1$.

Inductive step: assume the identity holds for some $n \geq 1$. We
show it holds for $n+1$. The left side becomes
\[
\sum_{i=1}^{n+1} i = \left(\sum_{i=1}^{n} i\right) + (n+1)
= \frac{n(n+1)}{2} + (n+1)
= \frac{n(n+1) + 2(n+1)}{2}
= \frac{(n+1)(n+2)}{2}.
\]
The right side, evaluated at $n+1$, is $\tfrac{(n+1)(n+2)}{2}$. The
two match. By induction, the identity holds for every $n \geq 1$.
\end{proof}

The proof has the canonical shape: a one-line base case and an
inductive step that uses the assumed identity to reduce the
$(n+1)$-case to a known $n$-case. The reader who works the proof
should notice that the inductive step relies on the algebraic
identity $n(n+1) + 2(n+1) = (n+1)(n+2)$, which is a factor-out of
$(n+1)$. Most inductive proofs end on an algebraic identity of this
kind; section 1.1's working library is the source.

\subsection{Worked example: sum of cubes}

A second worked induction shows a result an engineer would not guess
and that the induction confirms cleanly.

\begin{theorem}
For every natural number $n \geq 1$,
\[
\sum_{i=1}^{n} i^{3} = \left(\frac{n(n+1)}{2}\right)^{2}
= \left(\sum_{i=1}^{n} i\right)^{2}.
\]
\end{theorem}

\begin{proof}
Base case $n = 1$: the left side is $1$; the right side is
$(1 \cdot 2 / 2)^{2} = 1$. Both equal $1$.

Inductive step: assume $\sum_{i=1}^{n} i^{3} = \bigl(\tfrac{n(n+1)}
{2}\bigr)^{2}$. Add the $(n+1)$-th cube,
\[
\sum_{i=1}^{n+1} i^{3}
= \left(\frac{n(n+1)}{2}\right)^{2} + (n+1)^{3}
= (n+1)^{2}\left(\frac{n^{2}}{4} + (n+1)\right)
= (n+1)^{2} \cdot \frac{n^{2} + 4n + 4}{4}
= \left(\frac{(n+1)(n+2)}{2}\right)^{2}.
\]
This is the right side evaluated at $n+1$. By induction, the identity
holds for every $n \geq 1$.
\end{proof}

The result, that the sum of the first $n$ cubes equals the square of
the sum of the first $n$ integers, is the kind of identity an engineer
meets when summing a series and would not predict from the terms. The
induction does not explain why the identity is true in any deeper
sense; it certifies that it is. The factor-out of $(n+1)^{2}$ in the
inductive step is again an application of the section-1.1 working
library, which is the recurring pattern: an inductive step succeeds or
fails on an algebraic identity, and the algebra of section 1.1 is where
those identities live.

\subsection{Worked example: a recurrence with a closed form}

Induction also certifies a closed form for a recurrence, which is the
working case when an engineer has guessed a formula and needs to
confirm it. Consider the recurrence $T_{1} = 1$ and $T_{n} = 2 T_{n-1}
+ 1$ for $n \geq 2$, which counts, among other things, the moves in
the Tower of Hanoi puzzle and the worst-case nodes in a perfectly
unbalanced search.

\begin{theorem}
The recurrence $T_{1} = 1$, $T_{n} = 2 T_{n-1} + 1$ has closed form
$T_{n} = 2^{n} - 1$.
\end{theorem}

\begin{proof}
Base case $n = 1$: $T_{1} = 1 = 2^{1} - 1$. Inductive step: assume
$T_{n} = 2^{n} - 1$. Then $T_{n+1} = 2 T_{n} + 1 = 2(2^{n} - 1) + 1 =
2^{n+1} - 2 + 1 = 2^{n+1} - 1$. By induction the closed form holds for
every $n \geq 1$.
\end{proof}

The recurrence is solved here by guess-and-verify: the engineer guesses
$2^{n} - 1$ from the first few terms ($1, 3, 7, 15$) and induction
confirms the guess for all $n$. The guess-and-verify route is honest
engineering practice when the recurrence resists a direct solution
method, and induction is exactly the tool that converts a guess
checked on a few cases into a result proved on all. Chapter 4 of this
volume gives the systematic methods for solving recurrences without
guessing; induction remains the certificate that any candidate closed
form is correct.

\subsection{Worked example: induction on the cost of an algorithm}

The closed-form-by-induction move has a direct engineering use in
bounding the running time of a recursive algorithm. Merge sort splits an
array of $n$ elements into two halves, sorts each half recursively, and
merges the two sorted halves in linear time. Counting comparisons, the
cost obeys, for $n$ a power of two,
\[
T(1) = 0,
\qquad
T(n) = 2\,T(n/2) + n.
\]
The two recursive calls each cost $T(n/2)$, and the merge of two halves
of total length $n$ costs at most $n$ comparisons. We claim the closed
form $T(n) = n \log_{2} n$ and certify it by induction on $k$, where
$n = 2^{k}$.

\begin{theorem}
For $n = 2^{k}$ with $k \geq 0$, the recurrence $T(1) = 0$, $T(n) =
2\,T(n/2) + n$ has closed form $T(n) = n \log_{2} n$.
\end{theorem}

\begin{proof}
Base case $k = 0$: $n = 1$, and $T(1) = 0 = 1 \cdot \log_{2} 1$, since
$\log_{2} 1 = 0$. Both sides are zero.

Inductive step: assume the closed form holds at $n = 2^{k}$, that is,
$T(2^{k}) = 2^{k} \cdot k$. Evaluate the recurrence at $n = 2^{k+1}$,
whose half is $2^{k}$:
\[
T(2^{k+1}) = 2\,T(2^{k}) + 2^{k+1}
= 2 \cdot (2^{k} \cdot k) + 2^{k+1}
= 2^{k+1} k + 2^{k+1}
= 2^{k+1}(k + 1).
\]
The right side is $n \log_{2} n$ evaluated at $n = 2^{k+1}$, because
$\log_{2}(2^{k+1}) = k + 1$. By induction, $T(2^{k}) = 2^{k} \cdot k =
n \log_{2} n$ for every $k \geq 0$.
\end{proof}

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-recursion-tree.tex}

\Cref{fig:vol02:ch01:recursion-tree} draws the recursion tree the
induction certifies. Each level of the tree does $n$ units of merge
work in total, regardless of how the $n$ is partitioned across the
nodes at that level, and the tree has $\log_{2} n$ levels, so the total
is $n \log_{2} n$; the induction is the formal version of reading the
total off the tree. The result is the worst-case comparison count that
makes merge sort an $O(n \log n)$ algorithm, and the proof technique is
the one Volume VII uses throughout for algorithm analysis. The
restriction to powers of two is removed in the systematic treatment;
the closed form is correct in order of magnitude for every $n$, and the
induction here certifies the clean case from which the general bound
follows. The engineering reading is that an asymptotic running-time
claim is a claim over all input sizes, exactly the infinite-set claim
that no finite benchmark can establish and that induction can.

\subsection{Case study: a numerically stable running variance, derived
and verified}

We carry one algebra-and-proof thread from a stated need through a
derived closed form to an induction that certifies it, because the
discipline of the chapter is most visible when a single result is taken
all the way from motivation to certificate. The need is the one section
1.1's catastrophic-cancellation example exposed: an instrument streams
readings $x_{1}, x_{2}, \ldots$ and the engineer wants the running mean
and running variance after each reading, computed in one pass, without
storing the readings and without the cancellation that wrecks the
naive single-pass identity. The result is Welford's update, which the
engineer can write from memory once it is derived, and which this case
study derives and proves rather than quotes.

\paragraph{Step 1: the quantities and the goal.} After $n$ readings,
let the running mean be $\bar x_{n} = \frac{1}{n} \sum_{i=1}^{n} x_{i}$
and let the running sum of squared deviations be
\[
M_{n} = \sum_{i=1}^{n} (x_{i} - \bar x_{n})^{2},
\]
from which the sample variance is $s_{n}^{2} = M_{n}/(n - 1)$ for $n
\geq 2$. The goal is a pair of update rules that produce $\bar x_{n}$
and $M_{n}$ from $\bar x_{n-1}$, $M_{n-1}$, and the new reading $x_{n}$
alone, with every subtraction taken between quantities of comparable,
modest size.

\paragraph{Step 2: derive the mean update.} From the definition,
$n \bar x_{n} = (n - 1)\bar x_{n-1} + x_{n}$, so
\[
\bar x_{n} = \bar x_{n-1} + \frac{x_{n} - \bar x_{n-1}}{n}.
\]
The update adds a correction proportional to the residual $x_{n} -
\bar x_{n-1}$, the gap between the new reading and the old mean. The
subtraction is between two quantities of the same order as the data, so
no cancellation arises. We name the residual $\delta = x_{n} - \bar
x_{n-1}$ for the next step.

\paragraph{Step 3: derive the deviation-sum update.} The target update
is the identity
\[
M_{n} = M_{n-1} + (x_{n} - \bar x_{n-1})(x_{n} - \bar x_{n}),
\]
which we now prove is exactly the increment $M_{n} - M_{n-1}$. Write
$M_{n} - M_{n-1}$ by splitting off the new term and re-centring the old
ones,
\[
M_{n} - M_{n-1}
= \sum_{i=1}^{n} (x_{i} - \bar x_{n})^{2}
- \sum_{i=1}^{n-1} (x_{i} - \bar x_{n-1})^{2}.
\]
For each $i \leq n - 1$, write $x_{i} - \bar x_{n} = (x_{i} - \bar
x_{n-1}) - (\bar x_{n} - \bar x_{n-1})$ and abbreviate $c = \bar x_{n} -
\bar x_{n-1}$, the mean shift. Expanding the square,
\[
(x_{i} - \bar x_{n})^{2}
= (x_{i} - \bar x_{n-1})^{2}
- 2 c (x_{i} - \bar x_{n-1})
+ c^{2}.
\]
Summing over $i = 1, \ldots, n - 1$, the cross term vanishes because
$\sum_{i=1}^{n-1} (x_{i} - \bar x_{n-1}) = 0$ by the definition of
$\bar x_{n-1}$, leaving
\[
\sum_{i=1}^{n-1} (x_{i} - \bar x_{n})^{2}
= M_{n-1} + (n - 1) c^{2}.
\]
Adding back the new term $(x_{n} - \bar x_{n})^{2}$,
\[
M_{n} = M_{n-1} + (n - 1) c^{2} + (x_{n} - \bar x_{n})^{2}.
\]
From the mean update, $c = \bar x_{n} - \bar x_{n-1} = \delta/n$, so
$(n - 1) c^{2} = (n - 1)\delta^{2}/n^{2}$. Also $x_{n} - \bar x_{n} =
x_{n} - \bar x_{n-1} - c = \delta - \delta/n = \delta(n - 1)/n$, so
$(x_{n} - \bar x_{n})^{2} = \delta^{2}(n - 1)^{2}/n^{2}$. Summing the
two contributions,
\[
(n - 1) c^{2} + (x_{n} - \bar x_{n})^{2}
= \frac{\delta^{2}(n - 1)}{n^{2}}\bigl[1 + (n - 1)\bigr]
= \frac{\delta^{2}(n - 1)}{n}
= \delta \cdot \frac{\delta(n - 1)}{n}
= (x_{n} - \bar x_{n-1})(x_{n} - \bar x_{n}),
\]
which is the claimed increment. The deviation-sum update is established
as an exact algebraic identity, with no approximation.

\paragraph{Step 4: certify the algorithm by induction.} The update
rules define an algorithm: initialise $\bar x_{1} = x_{1}$, $M_{1} = 0$,
then for each new reading apply the mean update and the deviation-sum
update. We prove by induction that after processing $x_{1}, \ldots,
x_{n}$ the algorithm holds the true $\bar x_{n}$ and the true $M_{n}$.

\begin{theorem}
For every $n \geq 1$, the Welford recurrence produces $\bar x_{n} =
\frac{1}{n}\sum_{i=1}^{n} x_{i}$ and $M_{n} = \sum_{i=1}^{n} (x_{i} -
\bar x_{n})^{2}$.
\end{theorem}

\begin{proof}
Base case $n = 1$: the algorithm sets $\bar x_{1} = x_{1}$, which is the
mean of one reading, and $M_{1} = 0 = (x_{1} - \bar x_{1})^{2}$. Both
hold.

Inductive step: assume the algorithm holds the true $\bar x_{n-1}$ and
$M_{n-1}$ after $n - 1$ readings. The mean update computes $\bar
x_{n-1} + (x_{n} - \bar x_{n-1})/n$, which Step 2 showed equals the true
$\bar x_{n}$. The deviation-sum update computes $M_{n-1} + (x_{n} - \bar
x_{n-1})(x_{n} - \bar x_{n})$, which Step 3 showed equals the true
$M_{n}$. The inductive hypothesis supplied the true $\bar x_{n-1}$ and
$M_{n-1}$ that those updates consumed. Therefore the algorithm holds the
true $\bar x_{n}$ and $M_{n}$ after $n$ readings, and by induction for
every $n \geq 1$.
\end{proof}

The structure is the one the chapter has stressed. An algebraic identity
(Steps 2 and 3) is derived and checked; an induction (Step 4) lifts the
single-step identity to a claim over every $n$, which is the claim the
engineer actually needs, because the algorithm runs on streams of
unknown and unbounded length. The numerical payoff is the one section
1.1 set up: every subtraction in the updates is between a reading and a
running mean, quantities of the same modest order, so the catastrophic
cancellation of the naive single-pass identity never forms. The
reference implementation and a comparison against the naive sum-of-
squares form on a badly-scaled stream is in
\texttt{code/welford\_variance.py}, with the cancellation gap tabulated
in \texttt{data/welford-vs-naive.csv}. The thread from need to identity to
proof to code is the chapter's discipline applied end to end on one
result.

\subsection{Applied vignette: why a fixed-point iteration converges}

A second end-to-end thread joins an inequality, a geometric sum, and an
induction on one result an engineer relies on every time a solver is run:
the guarantee that a fixed-point iteration converges, and the bound on
how fast. Many engineering computations solve an equation $x = h(x)$ by
picking a start $x_{0}$ and iterating $x_{k+1} = h(x_{k})$, hoping the
sequence settles on a root. The steady-state filter of section 1.1's
neighbourhood was one instance; the Newton root-finder of Chapter 16 is
another; the load-flow solvers of Volume IX are a third. The question is
when the hope is justified, and the answer is an inequality lifted to all
$k$ by induction.

\paragraph{Step 1: the contraction hypothesis.} Suppose the map $h$ is a
\engterm{contraction}: there is a constant $L$ with $0 \leq L < 1$ such
that for every pair of points $a, b$ in the working interval,
\[
\abs{h(a) - h(b)} \leq L \, \abs{a - b}.
\]
The constant $L$ says that $h$ shrinks every distance by at least the
factor $L$. The reader who has met the derivative will recognise $L$ as a
bound on $\abs{h'}$; the reader who has not can take the inequality as
the hypothesis. Let $x^{\ast}$ be a fixed point, $x^{\ast} =
h(x^{\ast})$, whose existence we assume here and whose uniqueness the
contraction forces.

\paragraph{Step 2: one step shrinks the error.} Let $e_{k} = \abs{x_{k} -
x^{\ast}}$ be the error after $k$ steps. Apply the contraction with $a =
x_{k}$ and $b = x^{\ast}$, using $x_{k+1} = h(x_{k})$ and $x^{\ast} =
h(x^{\ast})$:
\[
e_{k+1} = \abs{h(x_{k}) - h(x^{\ast})} \leq L \, \abs{x_{k} - x^{\ast}}
= L \, e_{k}.
\]
One step multiplies the error by at most $L$. This is the single-step
inequality; the claim the engineer needs is about all steps at once.

\paragraph{Step 3: induction to all steps.} We claim $e_{k} \leq L^{k}
e_{0}$ for every $k \geq 0$, and prove it by induction. Base case $k = 0$:
$e_{0} \leq L^{0} e_{0} = e_{0}$, an equality. Inductive step: assume
$e_{k} \leq L^{k} e_{0}$. Then by Step 2 and the hypothesis,
\[
e_{k+1} \leq L \, e_{k} \leq L \cdot L^{k} e_{0} = L^{k+1} e_{0},
\]
where the middle inequality multiplies the inductive hypothesis by the
non-negative constant $L$, preserving direction. By induction the bound
holds for every $k$. Because $0 \leq L < 1$, the factor $L^{k} \to 0$, so
$e_{k} \to 0$ and the iteration converges to $x^{\ast}$ from any start in
the working interval. The convergence is geometric, the same $L^{k}$ tail
the geometric-series figure drew.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-cobweb-fixed-point.tex}

\Cref{fig:vol02:ch01:cobweb-fixed-point} draws the iteration as a
staircase between the map's curve and the diagonal, which is the picture
an engineer sketches to see at a glance whether a solver will settle. The
horizontal distance to the fixed point shrinks by the factor $L$ at every
step, so the staircase spirals inward when the slope has magnitude below
one and outward when it does not; the figure is the geometric statement
of the single-step inequality of Step 2.

\paragraph{Step 4: a stopping rule from the geometric sum.} The bound
also tells the engineer when to stop. The distance the iterate still has
to travel is bounded by summing the remaining steps, a geometric series:
\[
\abs{x_{k} - x^{\ast}}
\leq \sum_{j \geq k} \abs{x_{j+1} - x_{j}}
\leq \abs{x_{1} - x_{0}} \sum_{j \geq k} L^{j}
= \abs{x_{1} - x_{0}} \, \frac{L^{k}}{1 - L},
\]
which uses the infinite geometric sum with its $\abs{L} < 1$ convergence
condition, the same condition the contraction supplied. To reach an
absolute accuracy $\tau$, the engineer iterates until $\abs{x_{1} -
x_{0}}\, L^{k}/(1 - L) \leq \tau$, that is, until $k \geq \ln\!\bigl(\tau
(1 - L)/\abs{x_{1} - x_{0}}\bigr)/\ln L$. A contraction with $L = 0.5$
buys one decimal digit every three or four steps; a contraction with $L =
0.99$ crawls, needing hundreds of steps for the same digit, which is the
quantitative warning that a nearly-non-contracting map is a nearly-non-
convergent solver. The whole guarantee, existence of a limit, convergence
from any start, and a computable step count, rests on one inequality
lifted to all $k$ by induction and then summed as a geometric series.
Three of the chapter's threads carry one working result the engineer
uses whenever an iterative solver is trusted to terminate.

\begin{mastery}
The fixed-point vignette assumed a limit existed and proved the iterates
approach it. The assumption is discharged by the \engterm{Banach
fixed-point theorem}, which states that a contraction on a
\engterm{complete metric space} has exactly one fixed point, and the
iteration converges to it from every starting point. A metric space is a
set with a distance function $d(a, b)$ that is non-negative, zero only
when $a = b$, symmetric, and obeys the triangle inequality $d(a, c) \leq
d(a, b) + d(b, c)$, the same triangle inequality of section 1.2 promoted
to an axiom. The space is complete when every sequence whose terms
eventually crowd arbitrarily close together (a Cauchy sequence) actually
converges to a point of the space. The reals are complete; the rationals
are not, which is why a sequence of rational approximations to $\sqrt 2$
crowds together without a rational limit, the completeness counterpart of
section 1.3's irrationality proof. The theorem's proof is the vignette's
argument with the missing step supplied: the geometric bound shows the
iterates form a Cauchy sequence, completeness supplies the limit, and
continuity of the contraction shows the limit is the fixed point;
uniqueness follows because two fixed points would have to sit at distance
$d \leq L d$, forcing $d = 0$. The engineering reach is wide. The theorem
underwrites the convergence of Picard iteration for the ordinary
differential equations of Chapter 7, of successive-substitution load-flow
in power systems, and of the value-iteration algorithms of dynamic
programming. The working lesson is that an iterative method's guarantee
has three ingredients, a contraction constant below one, a distance that
obeys the triangle inequality, and a space in which limits exist, and an
engineer who trusts a solver has implicitly assumed all three.
\end{mastery}

\subsection{Applied vignette: the accumulated error of a long sum}

An engineer adding a great many numbers, the readings of a sensor array,
the terms of a series, the elements of a large data column, is running an
algorithm whose accuracy the chapter's tools predict before the code is
written. The addition $S_{n} = x_{1} + x_{2} + \cdots + x_{n}$ looks
exact, and over $\R$ it is; on the floating-point grid of section 1.6 it
is a chain of $n - 1$ roundings, and the question is how far the computed
sum $\tilde S_{n}$ can drift from the true one. The answer is an
inequality lifted to all $n$ by induction, and it motivates a better
algorithm the same reasoning then certifies.

\paragraph{Step 1: bound one rounding.} Each floating-point addition
returns the true sum perturbed by a relative error at most the unit
roundoff $u$: the computed value of $a + b$ is $(a + b)(1 + \delta)$ with
$\abs{\delta} \leq u$. This is the model the Ariane and Patriot cases of
section 1.7 fell foul of, stated as a bound rather than an anecdote.

\paragraph{Step 2: propagate through the running sum.} Add the terms
left to right, $\tilde S_{1} = x_{1}$ and $\tilde S_{k} = (\tilde
S_{k-1} + x_{k})(1 + \delta_{k})$ with $\abs{\delta_{k}} \leq u$. Each
partial sum inherits every earlier rounding and adds one of its own. We
claim, and prove by induction, that the total error is bounded to first
order by
\[
\abs{\tilde S_{n} - S_{n}} \leq (n - 1)\, u \sum_{i=1}^{n} \abs{x_{i}}
+ O(u^{2}),
\]
where the sum of magnitudes is the quantity the triangle inequality of
section 1.2 supplies. Base case $n = 1$: $\tilde S_{1} = x_{1} = S_{1}$,
error zero, and the bound reads $0 \leq 0$. Inductive step: assume the
bound at $n - 1$. The new partial sum is $\tilde S_{n} = (\tilde S_{n-1}
+ x_{n})(1 + \delta_{n})$, so to first order $\tilde S_{n} - S_{n} =
(\tilde S_{n-1} - S_{n-1}) + \delta_{n} S_{n} + O(u^{2})$. Taking
magnitudes and using the triangle inequality,
\[
\abs{\tilde S_{n} - S_{n}}
\leq \abs{\tilde S_{n-1} - S_{n-1}} + u \abs{S_{n}} + O(u^{2}).
\]
The inductive hypothesis bounds the first term by $(n - 2) u
\sum_{i=1}^{n-1} \abs{x_{i}}$, and $\abs{S_{n}} \leq \sum_{i=1}^{n}
\abs{x_{i}}$ by the triangle inequality, so the sum of the two is at most
$(n - 1) u \sum_{i=1}^{n} \abs{x_{i}}$ to first order, which is the bound
at $n$. By induction it holds for every $n$. The reading is that the
worst-case error of naive summation grows linearly in the number of
terms, the factor $(n - 1)$ out front, and it is worst when the terms
share a sign so that $\sum \abs{x_{i}}$ is as large as $\abs{S_{n}}$
allows. A sum of a million like-signed terms can lose six digits to this
alone.

\paragraph{Step 3: a better algorithm, and its bound.} Compensated
summation keeps a running correction $c$ that captures the low-order bits
each addition discards. At each step it forms $y = x_{k} - c$, then $t =
S + y$, then updates $c = (t - S) - y$ and $S = t$. The subtraction
$(t - S) - y$ recovers the part of $y$ that did not survive the addition
$S + y$, and feeds it into the next term. The error bound for compensated
summation is $\abs{\tilde S_{n} - S_{n}} \leq (2 u + O(u^{2}))
\sum_{i=1}^{n} \abs{x_{i}}$, with no factor of $n$ in the leading term.
The linear-in-$n$ growth is gone; the error is held near the single-
rounding floor regardless of how many terms are added.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-kahan-summation.tex}

\Cref{fig:vol02:ch01:kahan-summation} plots the two error curves against
the term count on logarithmic axes, with the $\sqrt{n}$ typical-case
curve between them. The naive error climbs along the $n u$ line the
induction bounded; the compensated error stays flat. The dashed
$\sqrt{n}$ reference is the average behaviour when the individual
roundings carry random signs and partly cancel, the same worst-case-
against-typical-case distinction the triangle-inequality figure drew for
error accumulation, here applied to the roundings of one long sum. The
engineering decision is the one the chapter keeps returning to: an
identity exact over $\R$, the associativity of addition that says the
order of summation does not matter, is only approximate on the machine,
and the algorithm that respects the approximation is not the one the
exact algebra suggested. The bound is what tells the engineer, before
running the code, whether the naive sum is good enough or whether the
compensated version is required; a thousand terms rarely need it, a
billion terms of mixed magnitude almost always do. Higham's treatment
gives the sharp constants and the variants \cite{text:higham-numerical};
the algebra-and-proof point is that the choice of summation algorithm is
an inequality-and-induction decision made on paper.

\subsection{Worked example: Bernoulli's inequality}

\begin{proof}[Proof of Bernoulli's inequality]
We show $(1 + x)^{n} \geq 1 + nx$ for $x \geq -1$ and $n \geq 0$.

Base case: $n = 0$. Both sides equal $1$.

Inductive step: assume $(1 + x)^{n} \geq 1 + nx$. Multiply both sides
by $(1+x)$, which is non-negative because $x \geq -1$. The
inequality direction is preserved:
\[
(1 + x)^{n+1} \geq (1 + nx)(1 + x) = 1 + (n+1) x + n x^{2}.
\]
$n x^{2} \geq 0$ for every real $x$ and every non-negative $n$, so
\[
(1 + x)^{n+1} \geq 1 + (n+1) x.
\]
By induction, the inequality holds for every $n \geq 0$.
\end{proof}

The condition $x \geq -1$ is required for the multiplication step:
multiplying by a negative quantity reverses the inequality. The
engineer who omits the condition has a proof that holds in
appearance and not in substance.

\subsection{Strong induction}

A variant called strong induction proves $P(n+1)$ from the assumption
that $P(k)$ holds for every $k$ with $1 \leq k \leq n$, rather than
only from $P(n)$. Strong induction is logically equivalent to
ordinary induction, but the form is more convenient when the
inductive step needs more than the immediately previous case.

\begin{theorem}
Every integer $n \geq 2$ is a product of one or more primes.
\end{theorem}

\begin{proof}
Base case: $n = 2$. $2$ is prime, hence a product of one prime.

Inductive step: assume every integer $k$ with $2 \leq k \leq n$ is
a product of primes. Consider $n+1$. If $n+1$ is prime, it is itself
a product of one prime. Otherwise $n+1 = a b$ for some integers $a$
and $b$ with $2 \leq a, b \leq n$. By the inductive hypothesis,
each of $a$ and $b$ is a product of primes. Their product is the
concatenation of the two factorisations, hence a product of primes.

By strong induction, every $n \geq 2$ is a product of primes.
\end{proof}

The use of strong induction is what licenses the appeal to the two
factors $a$ and $b$, both of which are smaller than $n+1$ but neither
of which is necessarily $n$ itself. Ordinary induction would not let
the writer make this step.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-strong-induction.tex}

\Cref{fig:vol02:ch01:strong-induction} contrasts the two dependency
structures. Ordinary induction advances one link at a time, each case
resting on its immediate predecessor. Strong induction lets a case rest
on the entire run before it, which is what the prime-factorisation proof
requires: the two factors $a$ and $b$ can sit anywhere below $n+1$, and
only an arrow reaching back to all earlier cases catches them. The two
forms are logically equivalent, and the choice between them is a matter
of which inductive hypothesis the step actually consumes.

\paragraph{Worked example: every postage above eleven cents, from two
stamps.} A second strong-induction case shows the pattern on a problem
an engineer recognises as a coin or denomination question. We claim that
every integer amount $n \geq 12$ can be made from $4$-cent and $5$-cent
stamps. The base cases are the four consecutive amounts $12 = 4 + 4 + 4$,
$13 = 4 + 4 + 5$, $14 = 4 + 5 + 5$, and $15 = 5 + 5 + 5$. For the
inductive step, take any $n \geq 16$ and assume every amount from $12$ up
to $n - 1$ is makeable. The amount $n - 4$ is at least $12$ and at most
$n - 1$, so by the inductive hypothesis it is makeable; adding one more
$4$-cent stamp makes $n$. The step reaches back four cases, not one,
which is why the proof needs four base cases and the strong form: an
ordinary induction resting on $n - 1$ alone could not step by four. The
number of base cases equals the reach of the step, and matching the two
is the discipline the all-horses paradox showed is easy to lose.

\subsection{Well-ordering and the minimal counterexample}

Induction has a mirror image that is often the cleaner tool for a
non-existence claim: instead of building the property up from a base
case, assume a counterexample exists and derive a contradiction from the
smallest one. The move rests on the \engterm{well-ordering principle},
that every non-empty set of natural numbers has a least element, and it
produces proofs that read as ``there is no smallest failure, so there is
no failure.''

\paragraph{Worked example: the same recurrence, by minimal
counterexample.} We reprove that the recurrence $T_{1} = 1$, $T_{n} = 2
T_{n-1} + 1$ has closed form $T_{n} = 2^{n} - 1$, this time without
climbing from the base case. Suppose the closed form fails for some $n$,
and let $S = \set*{n \geq 1 : T_{n} \neq 2^{n} - 1}$ be the set of all
$n$ where it fails. If the claim is false, $S$ is non-empty, so by
well-ordering $S$ has a least element $m$. Now $m \neq 1$, because $T_{1}
= 1 = 2^{1} - 1$ holds, so $m \geq 2$ and the recurrence applies:
$T_{m} = 2 T_{m-1} + 1$. Because $m$ is the least element of $S$, the
index $m - 1$ is not in $S$, so $T_{m-1} = 2^{m-1} - 1$ holds. Then
$T_{m} = 2(2^{m-1} - 1) + 1 = 2^{m} - 1$, which says $m \notin S$,
contradicting $m \in S$. The set $S$ has no least element, so by
well-ordering it is empty, and the closed form holds for every $n$. The
proof carries the same algebraic content as the induction of the
preceding subsection; the difference is the framing, which reaches for a
smallest failure rather than a first success. For a non-existence claim,
``no smallest counterexample'' is frequently the shortest route, and the
technique is the discrete form of the infinite-descent argument the
reader can recognise inside the $\sqrt 2$ proof, where an even $p$ and an
even $q$ would let the fraction be reduced forever, which well-ordering
forbids.

\begin{mastery}
The well-ordering principle and the principle of induction are
logically equivalent: each can be proved from the other, and either may
be taken as the axiom that distinguishes $\N$. Ordinary induction proves
well-ordering as follows. Suppose a non-empty set $S \subseteq \N$ had no
least element; let $P(n)$ be the statement ``no element of $S$ is less
than or equal to $n$.'' Then $P(0)$ holds (else $0$ would be a least
element), and if $P(n)$ holds then $n + 1 \notin S$ (else $n + 1$ would
be least, since nothing $\leq n$ is in $S$), so $P(n + 1)$ holds; by
induction $P(n)$ holds for all $n$, which forces $S$ empty, contradicting
non-emptiness. The reverse direction, well-ordering proving induction, is
the minimal-counterexample argument of the worked example above applied
to the set of $n$ for which $P(n)$ fails. The equivalence is why an
engineer may choose whichever framing fits the claim: a construction that
builds a property upward suits induction, and a claim that a failure
cannot occur suits the minimal-counterexample form. The same equivalence
extends, in the well-founded-induction box that closes this section, from
$\N$ to any order with no infinite descending chain, and the
well-ordering of $\N$ is the base case of that more general fact.
\end{mastery}

\subsection{Common errors in inductive proofs}

Three errors recur often enough to be worth naming.

\engterm{Missing the base case}. Without a base case, the inductive
step proves nothing. The famous parody is the proof that all horses
are the same colour, which assumes a base case at $n = 1$ and uses
the inductive step to extend to $n+1$, but the inductive step
silently assumes that two horses can be compared, which fails
exactly when the only horses available are the single horse of the
$n = 1$ case.

\engterm{The inductive step assuming the conclusion}. The writer
assumes $P(n+1)$ in the course of trying to prove $P(n+1)$ from
$P(n)$. This is circular and is sometimes hidden by an algebraic
manipulation that uses both the hypothesis and the conclusion as
substitutable.

\engterm{Variable confusion in the inductive step}. The inductive
step writes $P(n)$ as the hypothesis and proves $P(n+1)$. A common
slip is to use $n$ inside $P(n+1)$ where the value of $n$ is
ambiguous between the inductive variable and a fresh variable. The
discipline is to use distinct names: $n$ for the inductive variable
and, say, $k$ for any internal index inside the proof.

\subsection{The all-horses-same-colour paradox, worked}

The first error deserves a full worked treatment because the flaw is
subtle and instructive. The false ``theorem'' claims that in any set
of $n$ horses, all $n$ are the same colour. The base case $n = 1$ is
true: a single horse is the same colour as itself. The purported
inductive step takes a set of $n + 1$ horses, removes one to get a set
of $n$ (all the same colour by hypothesis), removes a different one to
get another set of $n$ (again all the same colour), and argues that
the two overlapping sets force all $n + 1$ horses to share a colour.

The flaw is in the overlap. The argument that the two sets of $n$
share a common horse, and therefore a common colour, requires the two
sets to actually overlap. For $n + 1 = 2$, the two sets of size
$n = 1$ are the single first horse and the single second horse, and
they do not overlap: there is no shared horse to chain the colours
through. The inductive step works for $n + 1 \geq 3$ but fails at the
single transition $n + 1 = 2$, and a chain of dominoes with one broken
link does not fall. The base case is at $n = 1$; the first inductive
step needed is $1 \to 2$; that step is exactly the one that fails. The
lesson is that an inductive step must hold for every $n$ from the base
case onward, and a step that holds for large $n$ but fails at the
smallest transition proves nothing. An engineer who tests an inductive
argument only at large $n$ can miss precisely this kind of boundary
failure.

\subsection{Structural and well-founded induction}\pathtag{mastery}

Ordinary induction climbs the natural numbers. The same principle
extends to any set equipped with a relation that has no infinite
descending chain, a structure called a \engterm{well-founded} order.
Structural induction is the version an engineer meets most often
outside arithmetic: to prove a property of every element of a
recursively defined structure (a tree, a list, a syntax of
expressions, a proof itself), prove it for the base constructors and
prove that each compound constructor preserves it. A property of every
binary tree is proved by checking the empty tree and checking that
attaching two subtrees with the property yields a tree with the
property. A property of every well-formed arithmetic expression is
proved by checking the literals and checking that each operator
preserves it.

\begin{mastery}
Well-founded induction is the general statement: if a property $P$
holds for an element $x$ whenever it holds for all elements below $x$
in a well-founded order, then $P$ holds for every element. Ordinary
induction is the case where the order is the usual order on $\N$ and
``below'' means ``smaller.'' Strong induction is the case where the
inductive hypothesis is taken over all smaller elements at once. The
well-foundedness condition, no infinite descending chain, is what
guarantees the induction terminates: a putative counterexample would
have a counterexample below it, and below that, forever, which a
well-founded order forbids. The engineering payoff is in reasoning
about programs that recurse on a structure: a recursive function
terminates exactly when its recursive calls descend in a well-founded
order, and the proof of termination is well-founded induction on that
order. The reader who writes recursive code is already using this
principle informally every time they argue that the recursion
``bottoms out''; the formal version names the order and checks that it
is well-founded. The same machinery, applied to the structure of
formal proofs, is how a proof assistant certifies that its own
proof-checking terminates.
\end{mastery}

\section{Functions, sets, mappings}\pathtag{standard}

Mathematics is, at one level, the disciplined manipulation of sets
and the functions that map them to one another. Engineers work with
functions constantly: a force as a function of position, a voltage
as a function of time, a temperature as a function of location. The
language of functions and sets gives the engineer the precision to
say which inputs produce which outputs, when the function is
invertible, when two functions compose, and what happens at the
boundary of the domain.

\subsection{Sets}

A set is an unordered collection of distinct objects. Two sets are
equal if and only if they contain the same elements. The empty set
$\emptyset$ has no elements. The standard set-builder notation
$\set*{x \in A : P(x)}$ denotes the subset of $A$ consisting of those
elements that satisfy the predicate $P$.

The standard sets the engineer will reach for are the natural numbers
$\N$, the integers $\Z$, the rationals $\Q$, the reals $\R$, and the
complex numbers $\C$. We will define $\R^{n}$ as the set of ordered
$n$-tuples of real numbers, the working space of vector calculus
(Chapter 8 onward).

The basic operations on sets are union ($A \cup B$), intersection
($A \cap B$), difference ($A \setminus B$), and Cartesian product
($A \times B = \set*{(a, b) : a \in A, b \in B}$). Each has a working
engineering use: union for the combined operating range of two
instruments, intersection for the common range, difference for the
extension of one over the other, Cartesian product for the joint
specification of two independent quantities.

The operations obey an algebra as disciplined as the algebra of numbers,
and two of its laws do load-bearing work in reliability and logic
design. De Morgan's laws relate complement to union and intersection,
\[
\overline{A \cup B} = \overline{A} \cap \overline{B},
\qquad
\overline{A \cap B} = \overline{A} \cup \overline{B},
\]
where the overline is complement within a fixed universal set. The laws
say that the complement of a union is the intersection of the
complements, and conversely; the operation flips when the complement
crosses it. Read as logic, with union as OR, intersection as AND, and
complement as NOT, De Morgan's laws are the identities NOT($P$ OR $Q$)
$=$ NOT $P$ AND NOT $Q$ and NOT($P$ AND $Q$) $=$ NOT $P$ OR NOT $Q$,
which are the transformations a digital designer applies to convert a
circuit between AND-OR and OR-AND form, and which a reliability engineer
applies to turn a statement about when a system succeeds into a statement
about when it fails. A system of two redundant components succeeds when
component one works OR component two works; by De Morgan it fails when
component one fails AND component two fails, which is the parallel-
redundancy calculation of Volume~I. Distributivity holds for sets as it
does for numbers, $A \cap (B \cup C) = (A \cap B) \cup (A \cap C)$, and
it is the algebra behind expanding a fault tree into its minimal cut
sets. The engineer who treats set operations as an algebra with stated
laws, rather than as a picture of overlapping circles, can manipulate a
specification symbolically and audit the result the same way an algebraic
identity is audited.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-de-morgan-redundancy.tex}

\Cref{fig:vol02:ch01:de-morgan-redundancy} draws De Morgan's law as the
reliability statement an engineer reaches for most often. The next
worked example carries the same law through a fault-tree calculation to
a number.

\paragraph{Worked example: a fault tree collapsed by De Morgan and
distributivity.} A safety interlock trips when a pump is running. The
pump runs when its power relay is closed AND its local disable switch is
open; call these events $R$ (relay closed) and $D$ (switch open), so the
hazardous state is $R \cap D$. A redundant guard is added: a second,
independent cutout $C$ that, when active, forces the interlock regardless
of the pump state. The system is safe when the guard is active OR the
pump is not in its hazardous state, that is, when $C \cup
\overline{(R \cap D)}$. The reliability engineer wants the failure event,
the complement of safety, because that is what the fault tree enumerates.
Taking the complement and applying De Morgan once,
\[
\overline{C \cup \overline{(R \cap D)}}
= \overline{C} \cap \overline{\overline{(R \cap D)}}
= \overline{C} \cap (R \cap D),
\]
where the double complement of $R \cap D$ collapses to $R \cap D$. The
failure event is $\overline{C} \cap R \cap D$: the guard is inactive AND
the relay is closed AND the switch is open, all three at once. This is a
single minimal cut set of three basic events, and it is exactly what the
engineer wants to read off, because a cut set names one combination that
alone causes failure. Distributivity does the second half of the work
when a guard has two independent channels $C = C_{1} \cup C_{2}$: the
failure event becomes $\overline{C_{1} \cup C_{2}} \cap R \cap D =
\overline{C_{1}} \cap \overline{C_{2}} \cap R \cap D$ by De Morgan again,
a four-event cut set, which correctly says a two-channel guard fails only
when both channels are inactive together with the hazardous pump state.
The algebra turned a specification stated on the safe side into the cut
sets stated on the failure side, symbolically and auditably, with no
picture of overlapping circles and no appeal to intuition about which
events conspire.

\subsection{Functions}

A function $f$ from a set $A$ to a set $B$, written $f : A \to B$,
assigns to each element $a \in A$ exactly one element $f(a) \in B$.
$A$ is the domain and $B$ is the codomain. The image of $f$ is the
subset $\set*{f(a) : a \in A}$ of $B$, which may be smaller than $B$.

Three properties classify functions for working purposes.

\engterm{Injective} (one-to-one): different inputs produce different
outputs. $f(a_{1}) = f(a_{2})$ implies $a_{1} = a_{2}$.

\engterm{Surjective} (onto): every element of the codomain is the
image of some element of the domain. The image equals the codomain.

\engterm{Bijective}: both injective and surjective. A bijection has
a two-sided inverse $f^{-1} : B \to A$, characterised by $f^{-1}(f(a))
= a$ and $f(f^{-1}(b)) = b$.

The square function $f(x) = x^{2}$ is not injective on $\R$, because
$f(2) = f(-2)$. It is injective on $[0, \infty)$. The discipline of
restricting the domain to make a function injective is exactly the
discipline that lets the engineer define $\sqrt{x}$ as the unique
non-negative real number whose square is $x$. The square root, like
many engineering inverses, is a partial inverse, defined only on the
restricted domain on which the original function is bijective.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-function-classification.tex}

\Cref{fig:vol02:ch01:function-classes} shows the three classes on
small finite sets. The engineering uses are direct.
A sensor calibration is bijective on its working range, with a
two-sided inverse that converts indicated value back to true value.
A many-to-one classifier (an analog-to-digital converter mapping a
continuous voltage to one of $2^{N}$ codes) is surjective onto the
code set and not injective; the loss of information is exactly the
quantisation error of Volume I Chapter 6. A measurement function
whose codomain includes states the instrument cannot actually
produce (a strain-gauge bridge that nominally outputs any voltage
but actually saturates) is injective on its working domain and not
surjective onto the nominal codomain; the unreachable codomain is
where saturation, clipping, and out-of-range diagnostics live.

\paragraph{Worked example: the logistic function as a partial
inverse.} The logistic map $\sigma(x) = 1/(1 + e^{-x})$ takes
$\R \to (0, 1)$ and is bijective from $\R$ onto $(0, 1)$. Its
inverse, the logit $\text{logit}(p) = \log(p / (1 - p))$, is
defined on the open interval $(0, 1)$ and undefined at $p = 0$ or
$p = 1$. The engineer who applies a logit transform to a measured
probability close to 0 or 1 needs to know which inverse is being
applied and where it ceases to be defined; clipping the input to
$[\epsilon, 1 - \epsilon]$ for a small $\epsilon$ is the working
discipline, and the choice of $\epsilon$ becomes the
domain-restriction step that turns a partial inverse into a useful
one.

\subsection{Composition}

The composition of $g : B \to C$ with $f : A \to B$ is the function
$(g \circ f) : A \to C$ defined by $(g \circ f)(a) = g(f(a))$. The
composition is the engineering picture of two systems in series: the
output of the first is the input of the second. Composition is
associative, $h \circ (g \circ f) = (h \circ g) \circ f$, but not
commutative; $f \circ g$ and $g \circ f$ are different functions in
general.

The composition of two injective functions is injective. The
composition of two surjective functions is surjective. The
composition of two bijections is a bijection, with inverse $(g \circ
f)^{-1} = f^{-1} \circ g^{-1}$. The reversal of order is the same
reversal that appears in matrix transposition $(AB)\transpose =
B\transpose A\transpose$ in Chapter 8.

\paragraph{Worked example: the order of undo is the reverse of do.} A
measurement pipeline first subtracts a zero offset, $f(v) = v - v_{0}$,
then applies a gain, $g(w) = k w$, so the calibrated reading is $(g \circ
f)(v) = k(v - v_{0})$. To recover the raw voltage from a calibrated
reading, the engineer inverts. The inverse of the composition reverses
the order: $(g \circ f)^{-1} = f^{-1} \circ g^{-1}$. Concretely,
$g^{-1}(y) = y/k$ undoes the gain first, then $f^{-1}(w) = w + v_{0}$
adds the offset back, giving $(f^{-1} \circ g^{-1})(y) = y/k + v_{0}$.
Checking against the forward map, $(f^{-1} \circ g^{-1})(k(v - v_{0})) =
k(v - v_{0})/k + v_{0} = v$, the raw voltage returns. Applying the
inverses in the forward order instead, $g^{-1}$ then $f^{-1}$ read
left-to-right as $f^{-1}$ first, would add the offset to a still-scaled
value and get the wrong answer. The reversal is not a convention; it is
forced, the same way the last item packed is the first unpacked. An
engineer who chains transforms (offset, gain, coordinate rotation,
unit conversion) inverts them in strict reverse order, and the composition
identity is the statement of that rule.

\subsection{Relations and equivalence}

A function is a special kind of relation. A \engterm{relation} on sets
$A$ and $B$ is any subset of the Cartesian product $A \times B$; a
function is the case where each $a \in A$ appears in exactly one pair.
A relation on a single set $A$ (a subset of $A \times A$) can carry
three properties whose combination an engineer uses constantly. A
relation $\sim$ is \engterm{reflexive} if $a \sim a$ for every $a$,
\engterm{symmetric} if $a \sim b$ implies $b \sim a$, and
\engterm{transitive} if $a \sim b$ and $b \sim c$ imply $a \sim c$. A
relation with all three is an \engterm{equivalence relation}, and it
partitions $A$ into disjoint \engterm{equivalence classes}, each the
set of elements mutually related.

The engineering uses are direct and frequent. Congruence mod $n$ is
the equivalence relation behind every cyclic counter, every checksum,
and the modular arithmetic of Volume VII. Dimensional homogeneity is
an equivalence relation on physical quantities: two quantities are
related when they share dimensions, and the classes are the
dimensional groups the engineer balances at every step. Tolerance
binning sorts manufactured parts into equivalence classes by measured
value, with each bin an equivalence class under the relation ``within
the same tolerance band.'' Naming a relation as an equivalence
relation is what licenses the engineer to treat the whole class as a
single object, which is the compression that makes modular arithmetic
and dimensional analysis tractable.

\paragraph{Worked example: a relation that is not an equivalence.}
The relation ``$a$ and $b$ agree to within $1\,\si{\milli\meter}$'' on
a set of length measurements is reflexive and symmetric but not
transitive: $a$ within $1\,\si{\milli\meter}$ of $b$ and $b$ within
$1\,\si{\milli\meter}$ of $c$ allows $a$ and $c$ to differ by up to
$2\,\si{\milli\meter}$. The failure of transitivity is the formal
reason that ``close enough'' does not partition a set into clean
classes, and it is the reason a tolerance scheme must bin against a
fixed reference rather than against neighbouring parts. An engineer who
chains pairwise ``close enough'' comparisons and expects a consistent
grouping has assumed transitivity that the relation does not have.

\paragraph{Worked example: casting out nines as an equivalence-class
check.} Congruence mod $9$ is an equivalence relation on the integers:
$a \sim b$ when $9$ divides $a - b$. It is reflexive, symmetric, and
transitive, so it partitions the integers into nine classes, and the
class of any integer is its remainder on division by $9$. The class of a
base-ten number equals the class of its digit sum, because $10 \equiv 1
\pmod 9$ makes every power of ten congruent to $1$, so $\sum_{k} d_{k}
10^{k} \equiv \sum_{k} d_{k} \pmod 9$. This licenses a one-line
arithmetic check. To test $2738 \times 41 = 112{,}258$, reduce each
factor to its class: $2 + 7 + 3 + 8 = 20 \to 2 + 0 = 2$, and $4 + 1 =
5$. The product of the classes is $2 \times 5 = 10 \to 1$. The claimed
answer reduces to $1 + 1 + 2 + 2 + 5 + 8 = 19 \to 1 + 9 = 10 \to 1$. The
two agree, so the answer passes. Had a digit been transposed, the classes
would disagree with high probability and the error would surface without
redoing the multiplication. The check works because congruence mod $9$ is
compatible with addition and multiplication: operating on the whole class
gives the same class as operating on representatives, which is exactly
the compression that naming a relation an equivalence relation buys. The
check cannot catch every error, a transposition of two digits leaves the
digit sum unchanged, so it is a necessary and not a sufficient test, the
same balance-test caveat the partial-fraction check carried in section
1.1.

\subsection{Restriction, extension, image, preimage}

The restriction $f \restriction_{A'}$ of $f : A \to B$ to a subset
$A' \subseteq A$ is the function $A' \to B$ defined by
$f \restriction_{A'}(a) = f(a)$. The image of a subset $A' \subseteq
A$ is $f(A') = \set*{f(a) : a \in A'}$. The preimage of a subset
$B' \subseteq B$ is $f^{-1}(B') = \set*{a \in A : f(a) \in B'}$.
The preimage exists for every function, even ones that are not
invertible; the notation $f^{-1}$ is overloaded between the inverse
function (when $f$ is bijective) and the preimage (always defined),
and the engineer must know from context which is meant. Section 1.6
takes up notation overloading as a discipline issue.

The asymmetry between image and preimage is worth stating precisely,
because it is the source of a recurring error. The preimage respects
unions, intersections, and complements exactly: $f^{-1}(B_{1} \cup
B_{2}) = f^{-1}(B_{1}) \cup f^{-1}(B_{2})$, and the same clean equality
holds for intersection and for complement. The forward image does not.
The image respects unions, $f(A_{1} \cup A_{2}) = f(A_{1}) \cup
f(A_{2})$, but for intersection it gives only containment, $f(A_{1} \cap
A_{2}) \subseteq f(A_{1}) \cap f(A_{2})$, and the containment can be
strict. A one-line witness: for $f(x) = x^{2}$ with $A_{1} = \{-1\}$ and
$A_{2} = \{1\}$, the sets are disjoint so $f(A_{1} \cap A_{2}) =
f(\emptyset) = \emptyset$, yet $f(A_{1}) \cap f(A_{2}) = \{1\} \cap \{1\}
= \{1\}$. The two functions collided the two inputs onto one output, and
the forward image cannot see that the two came from disjoint sets. The
working consequence is that an engineer who reasons about which outputs a
system can produce should push the reasoning through the preimage
whenever possible, because the preimage preserves the set algebra and the
image does not. The preimage is also the correct tool for a
specification stated on outputs: the set of inputs that meet an output
tolerance $B'$ is exactly the preimage $f^{-1}(B')$, defined whether or
not $f$ is invertible, and computing it does not require the inverse
function to exist.

\subsection{Functions in engineering}

Three working classes of function recur often enough to name. Linear
functions $f(x) = mx + b$ are the simplest. Polynomial functions
extend the working library; the chapter introduction introduced
their properties. Rational functions, ratios of polynomials, are
the next level of complexity and appear in transfer functions,
filter design, and asymptotic analysis. Exponentials and logarithms
are introduced formally in Chapter 2 of this volume; trigonometric
functions in Chapter 3. The working point here is that the function
concept covers all of these, and that the discipline of stating the
domain explicitly is the discipline that prevents an engineer from
applying $\log$ to a negative number, $\sqrt{\cdot}$ to a negative
real, or $\arcsin$ to a value outside $[-1, 1]$.

A point the definition makes and casual usage blurs is that a function is
not the same object as a formula. A function is a domain, a codomain, and
a rule assigning one output to each input; a formula is one possible
rule, and the same function can wear several formulas. The function that
sends $x$ to $x$ on the reals is also written $\sqrt{x^{2}}$ only on the
non-negative reals, where the two formulas agree; on the negatives
$\sqrt{x^{2}} = -x$, so the single formula $\sqrt{x^{2}}$ is the
function $\abs{x}$, not the identity. The discipline this forces is the
check of \engterm{well-definedness}: whenever a rule offers more than one
route to the output, the routes must agree, or the rule is not a function
at all. The check bites hardest when the input is specified indirectly.
A rule that assigns to a fraction $p/q$ the value $p + q$ is not
well-defined, because $1/2$ and $2/4$ are the same input yet give $3$ and
$6$; the rule depends on the representative, not the number. A rule that
assigns to an angle its cosine is well-defined even though $\theta$ and
$\theta + 2\pi$ are different representatives of the same direction,
because cosine returns the same value on both. The engineering setting
where this recurs is any quantity defined on equivalence classes: a
function of a physical direction must return the same value on every
angle that names that direction, and a function of a residue class must
return the same value on every integer in the class. Naming the domain,
and checking that the rule respects whatever identifications the domain
carries, is the difference between a function and a formula that only
looks like one.

\subsection{Worked example: when an inverse exists}

A temperature sensor outputs a voltage $V = f(T)$ that the engineer
must invert to recover the temperature $T$ from a measured voltage. The
inverse exists, as a function, exactly when $f$ is injective on the
working temperature range. A thermocouple's voltage is a smooth,
strictly increasing function of temperature over its rated span, so
$f$ is injective there and $f^{-1}$ is well defined; the engineer
inverts a calibration curve and recovers $T$ uniquely. A thermistor's
resistance is strictly monotone in temperature over its span, so the
same holds. The failure case is a sensor whose output is not monotone:
a bare silicon diode used as a temperature sensor has a response that
turns over at low temperature, so two distinct temperatures can produce
the same voltage, and $f$ is not injective across the turnover. On the
non-monotone span the inverse is not a function; the engineer must
restrict the domain to one monotone branch, exactly the
domain-restriction discipline that defines $\sqrt{\cdot}$ from
$x \mapsto x^{2}$. The working rule is that an instrument is usable as
a measurement only on a domain where its response function is
injective, and identifying that domain is part of specifying the
instrument.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-monotone-inversion.tex}

\Cref{fig:vol02:ch01:monotone-inversion} draws the two cases side by
side. The monotone response admits a calibration inverse over its whole
span; the response that turns over shares one output between two
temperatures, and on the span that includes the turnover no inverse
function exists. The figure is the pictorial statement of the working
rule: an instrument is usable as a measurement only on a domain where its
response is injective.

\begin{mastery}
The injective-surjective-bijective vocabulary scales from finite sets
to infinite ones, and the scaling reveals that infinite sets come in
different sizes. Two sets have the same \engterm{cardinality} when a
bijection exists between them. The natural numbers, the integers, and
the rationals all have the same cardinality, called countable: each can
be put in bijection with $\N$, even though $\Z$ and $\Q$ appear larger.
The reals are not countable; Cantor's diagonal argument (the mastery
box of section 1.3) constructs a real missing from any proposed
bijection with $\N$, so no such bijection exists, and the reals are
strictly larger. The working consequence for an engineer is in
computation and representation: a countable set can be enumerated by a
program, one element after another, with every element reached in
finite time; an uncountable set cannot. The real numbers an engineer
manipulates are always a countable, finite subset, the representable
floating-point values, and the gap between the uncountable continuum of
the model and the finite grid of the machine is the gap that section
1.7's Patriot and Ariane cases fell into. The cardinality vocabulary
gives the precise statement of why the model and the machine can never
fully coincide: the model's domain is uncountable, the machine's is
finite, and no bijection bridges them.
\end{mastery}

\subsection{The role of mappings in later chapters}

We close this section with a brief preview. Volume II Chapter 8
introduces linear maps as functions from $\R^{n}$ to $\R^{m}$ that
preserve the vector-space operations. Volume III Chapter 4
introduces the gradient as a function from a scalar field to a
vector field. Volume VII Chapter 1 introduces algorithms as
functions from input strings to output strings, with running time
and space as functions of input size. In each, the working concept
of function from this section is the foundation. The reader who
knows what an injective function is in $\R$ knows what an injective
linear map is in $\R^{n}$.

\section{Notation discipline: how engineers should write
mathematics}\pathtag{core}

A reader who can do the algebra and the proofs of the previous
sections still produces unreadable mathematics if their notation is
disorganised. The notation is the contract between the writer and
the reader: it tells the reader which symbol means what, in which
domain it lives, in which units it carries, and how the symbol
relates to the others on the page. We list the discipline as a set
of rules each of which has been violated, repeatedly, by working
engineers, and each of which has at one time or another been
implicated in a real failure.

\subsection{Definition before use}

Every symbol that appears in a calculation is defined before its
first use. The definition states the symbol's meaning, its domain
(integer, real, complex, vector, matrix, function), its units when
relevant, and the symbol it should not be confused with.

A common failure is to introduce $\rho$ as ``the density'' without
stating which density. The chapter on heat transfer means mass
density per unit volume; the chapter on probability means a
probability density function; the chapter on statistical inference
means a correlation coefficient. The same Greek letter has at least
three working meanings. The engineer who writes a paper with $\rho$
appearing in three sections must define it three times or, better,
use three different symbols.

\subsection{Units in every quantity}

A quantity has a unit. The unit is part of the quantity. An equation
that mixes units without explicit conversion is wrong before it is
solved. We carry units through every calculation, and we audit at
each interface. The discipline is the same as the unit discipline of
Volume I Chapter 2, applied to algebraic notation rather than
measurement. The Mars Climate Orbiter case (section 1.7) shows what
happens when the discipline lapses at an interface.

\subsection{Indices typeset properly}

A subscript is a label that selects a member of a sequence or a
component of a vector. A superscript is most commonly an exponent.
Confusing the two is a recurring source of error, especially when
exponents appear next to subscripts. The standard form $a_{i}^{2}$
means the $i$-th member of the sequence $a$, squared. $a^{2}_{i}$
means the same thing but is harder to read. $a^{i}$ for the $i$-th
member is acceptable in tensor notation (where the position of the
index carries information about transformation behaviour) but not
in ordinary notation, where it would be read as ``$a$ raised to the
$i$.''

The engineer who writes $a_{i}^{2}$ and means ``the $i$-th member
squared'' should also state, on first use, that the convention is
member-then-square. The reader who reads the convention as
``the square's $i$-th component'' is reading the formula correctly
under a different convention. The notational risk is real and
documented; the discipline is to state the convention.

\subsection{Avoiding overloaded symbols}

A symbol is overloaded when it has two distinct meanings in the
same document. Overloading is sometimes unavoidable; $f$ may be a
function and $f$ may be a frequency in the same volume. The
discipline is to name the overload explicitly when it is introduced
and to use distinct fonts, subscripts, or accents to keep the two
apart. $f(t)$ for the function and $f_{0}$ for the frequency, with a
sentence stating the two distinct uses, is acceptable; $f$ standing
in for both without comment is a notational hazard.

The variables $i$, $j$, $k$ deserve special note. They are commonly
used as integer indices, as imaginary units (especially $i$ and
$j$), and as Cartesian unit vectors ($\hat{\imath}$, $\hat{\jmath}$,
$\hat{k}$). The engineer who uses $i$ as both an index and the
imaginary unit, in the same calculation, is asking for an error.
Electrical engineering's convention is to use $j$ for the imaginary
unit because $i$ is current; mathematics' convention is the
opposite. The reader who carries notation between two communities
must state which convention applies in which section.

\subsection{Equations numbered and cross-referenced}

In a paper or chapter long enough to refer back to its own
equations, the equations are numbered and the cross-references use
the numbers. The discipline lets a reader who returns to an equation
forty pages later identify it without ambiguity. LaTeX's equation
environment provides this; the engineer's job is to use it.

A common failure is to write ``as we have already shown'' without a
cross-reference. The reader who has not memorised the chapter cannot
find the prior result and cannot verify the current step. The cost
of inserting a label and a cross-reference is one minute; the cost
to the reader of not having one is much larger.

\subsection{The notation table}

A document with more than a handful of symbols benefits from a
notation table. The table lists every symbol, its meaning, its
domain or type, its units, and the section where it is first
introduced. Volume I introduced this discipline through its
\repopath{frontmatter/notation.tex} file, which the reader may use
as a template. A working notation table catches overloading early:
the writer who tries to add $\rho$ to the table and finds it is
already there has noticed the overload before the reader does.

\subsection{The interface as the place where notation crosses}

Most notation failures are interface failures. The Mars Climate
Orbiter team that produced impulse data in pound-force seconds was
internally consistent. The team that consumed the data in newton
seconds was internally consistent. The interface between them was
the place where the notation crossed; the conversion was missing.
We will name this pattern through the rest of the volume. Every
interface is a place where the notation must be audited. Every
notation table is an audit document. The discipline of writing
mathematics for an engineering reader is, in large part, the
discipline of making the interfaces explicit.

\begin{archetype}[Interface]
Notation is a contract at an interface. A symbol carries meaning only
because the writer and the reader have agreed, explicitly or
implicitly, on its domain, units, precision, range, and provenance. The
interface is the place where one party's representation meets another
party's representation, and it is the place where the agreement must be
made explicit and audited. The archetype recurs across the entire book.
An application programming interface is a contract on the type and
range of every argument. A mechanical drawing is a contract on
dimension, tolerance, and datum. A data sheet is a contract on the
operating conditions under which a stated specification holds. A
standard is a contract on the meaning of a measured quantity across
every laboratory that adopts it. Each is an interface where
representation crosses, and each fails the same way notation fails: a
quantity is produced under one agreement and consumed under another,
with no audit at the boundary. The Mars Climate Orbiter, Air Canada
Flight 143, Ariane 5, Hubble, and Patriot cases of section 1.7 are five
realisations of one archetype. The interface archetype is the
counterpart, in the world of communication between systems, of the
balance archetype in the world of a single expression: balance keeps
one party's algebra honest, and the interface contract keeps two
parties' algebra compatible.
\end{archetype}

\subsection{Representation: precision, range, and the floating-point
grid}

Units are the most visible part of a representation, and they are not
the whole of it. A quantity stored in a computer carries a precision
(how many significant figures the format holds) and a range (the
largest and smallest magnitudes it can represent). The Ariane 5 and
Patriot cases of section 1.7 are representation failures of exactly
this kind: a value valid in $\R$ overflowed a 16-bit integer, and a
decimal $0.1$ had no exact binary representation. The discipline of
notation extends to stating, at every interface that crosses into
finite-precision arithmetic, the precision and range the consuming
system assumes.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-floating-point-line.tex}

\Cref{fig:vol02:ch01:floating-point-line} shows the structural fact an
engineer must carry: representable floating-point numbers are not
evenly spaced on the real line. Within each interval between
consecutive powers of two the spacing is constant, and it doubles at
every power of two. An algebraic identity that is exact over the
unbounded, infinitely divisible reals is only approximate over this
grid, and the approximation error accumulates in ways the exact algebra
does not predict. Goldberg's survey \cite{text:goldberg1991-floating-point}
and Overton's text \cite{text:overton-ieee754} are the working
references; the engineering point for this chapter is that
``definition before use'' includes the definition of the number format,
and that an interface specification silent on precision and range has
left a contract term blank.

\subsection{Worked example: a notation rewrite, before and after}

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-notation-rewrite.tex}

\Cref{fig:vol02:ch01:notation-rewrite} sets a defective fragment beside
its corrected form, which is the operation the project of section 1.8
asks the reader to perform on a real published page. The left fragment
states Fourier's law of heat conduction as $q = k\,A\,dT/dx$, uses
$\rho$ for thermal diffusivity, leaves $c$ undefined, attaches no
units, and writes a node temperature as $T_{i}^{2}$. Six defects sit
in four lines. The sign on Fourier's law is dropped: heat flows down
the temperature gradient, so the law is $q = -k\,A\,dT/dx$, and the
missing sign would invert the predicted flow direction. The symbol
$\rho$ is overloaded, standing for mass density in most of the
hypothetical paper and for diffusivity here; the fix renames the
diffusivity $\alpha$ and writes it as $\alpha = k/(\rho c_{p})$ with
$\rho$ restored to density. The symbol $c$ is used without definition;
the fix names it $c_{p}$, the specific heat at constant pressure, with
its units. No quantity carries units; the fix attaches
$\si{\watt\per\meter\per\kelvin}$ to $k$, $\si{\meter\squared}$ to $A$,
$\si{\kelvin}$ to $T$, and $\si{\meter\squared\per\second}$ to
$\alpha$. The node temperature $T_{i}^{2}$ is ambiguous between the
$i$-th node's temperature squared and the square's $i$-th component;
the fix writes $(T_{i})^{2}$ and states the index range $i = 1, \ldots,
N$. The governing law carries no cross-reference; the fix points to the
section that introduces it. The corrected fragment is longer by a few
words and unambiguous to a reader who has not seen the original, which
is the standard the chapter holds.

\subsection{A short style guide}

We close with a list of working rules an engineer can apply at
writing time.

\begin{itemize}[leftmargin=*]
  \item State the domain of each variable on first use. ``Let $x \in
    \R$'' is one syllable longer than ``Let $x$,'' and it eliminates
    a class of error.
  \item Carry units through every step. If the calculation collapses
    into a pure number, state where and why.
  \item Use a distinct font (Roman for units, italic for variables,
    bold for vectors, sans-serif for tensors) and use it consistently.
  \item Reserve $i$, $j$, $k$ for indices in ordinary algebra.
    Use $\sqrt{-1}$ explicitly or write the imaginary unit in upright
    Roman as $\mathrm{i}$ when introducing it for the first time.
  \item Number every equation that you might refer back to. The cost
    of an unused number is small; the cost of a missing one is
    large.
  \item Write the notation table first or last; do not write it in
    the middle. The first version anchors the writer's discipline;
    the last version audits the writer's lapses. Both are useful.
\end{itemize}

The reader who applies these rules to the project of section 1.8
will produce a rewrite of a poorly-typeset paper that is recognisably
better than the original. The discipline is a habit, learned by
practice, and reinforced by the discomfort of reading one's own
mathematics six months later and not knowing what $\alpha$ meant.

\section{Failure: notation collapse and the cost of imprecision}\pathtag{core}

Every chapter in this work has a failure section. The chapter on
algebra and proof discipline closes on the failure mode the chapter's
discipline exists to prevent: notation collapse at an engineering
interface. The chronic version is a paper unreadable to its second
audience; the acute version is the loss of a spacecraft.

\subsection{The Mars Climate Orbiter, 1999}

NASA's Mars Climate Orbiter was lost on 23 September 1999 when ground
software supplied small-forces impulse data in pound-force seconds
while the navigation interface required newton seconds, producing a
cumulative trajectory error that brought the spacecraft into the
Martian atmosphere during orbital insertion \cite{acc:nasa-mco-mib}.

\paragraph{Technical mechanism.}
The Mishap Investigation Board found that the ground software
program \texttt{SM\_FORCES} produced thruster-generated
change-in-velocity data using thruster performance values in English
units (pound-force seconds). The interface specification with the
navigation team's \texttt{AMD} file required impulse data in metric
units (newton seconds). The conversion factor of approximately
$4.45$ between the two never appeared in any handoff. Over the
roughly nine-month cruise, the small per-burn impulse error
accumulated through angular-momentum-desaturation manoeuvres and
trajectory-correction manoeuvres into a cumulative trajectory
deflection. At Mars arrival, the spacecraft's actual closest
approach was approximately $57 \,\si{\kilo\meter}$ above the Martian
surface, well below the planned $193 \,\si{\kilo\meter}$ closest
approach and below the spacecraft's atmospheric-survival altitude.
The spacecraft entered the Martian atmosphere and was destroyed
\cite{acc:nasa-mco-mib}. The technical failure sat at the interface
between contractor data and the navigation model. Each side of the
interface was internally consistent. The interface document
specified metric units. The producing software did not enforce the
specification, and the consuming software did not verify it. A
representative audit table that, had it been required at the
interface, would have caught the slip is shown in
\texttt{data/mco-unit-handoff.csv}; the table's structure is
trivial, and the discipline cost of populating it would have been
trivial against the cost of the lost spacecraft.

\paragraph{Organisational context.}
The Mishap Investigation Board's recommendations centred on missing
barriers rather than on individual error. The board found inadequate
verification of unit consistency across the contractor (Lockheed
Martin Astronautics) and operator (Jet Propulsion Laboratory)
interface, inadequate review of small-forces data delivered between
the two organisations, inadequate testing against flight data during
cruise, and inadequate independent review of the trajectory-
correction manoeuvre process. The board did not blame any
individual; the board's framing was that the process should not have
allowed the error to reach Mars, and that several places existed
where audit, review, or test would have caught the discrepancy
\cite{acc:nasa-mco-mib}. The contractor-operator split is the
organisational shape that makes interface failures recur: each
organisation's internal discipline is high, and each organisation's
visibility into the other's discipline is low.

\paragraph{Lesson for algebraic and proof discipline.}
The interface is the place where one party's algebra meets another
party's algebra. Inside Lockheed Martin's \texttt{SM\_FORCES} the
algebra balanced: impulse in pound-force seconds, integrated against
thruster firing time in seconds, yields change in velocity in the
internal units. Inside JPL's \texttt{AMD} navigation code the algebra
also balanced: impulse in newton seconds, integrated against
spacecraft mass in kilograms, yields change in velocity in metres
per second. The interface variable carried no dimensional tag; the
consuming code applied its own assumed dimension; the conversion
factor of $4.45$ was the missing multiplication that would have
restored the balance. The chapter's principle reads this case as the
canonical reminder that an algebraic identity holds only when the
underlying quantities live in the units the identity assumes.
\Cref{fig:vol02:ch01:mco-interface} draws the cascade from the
unaudited interface through the cumulative error path to mission
loss.

\input{volumes/02-form/ch01-algebra-and-proof/figures/fig-mco-interface.tex}

\subsection{Air Canada Flight 143, 1983}

A second case shows the same failure mode at human scale. On 23 July
1983 a Boeing 767, Air Canada Flight 143, ran out of fuel at
$41{,}000\,\text{ft}$ because the fuelling crew used a density
factor near $1.77 \,\text{lb/L}$ where the calculation required
about $0.80 \,\si{\kilogram\per\liter}$, departing with about half
the fuel its flight plan required; the aircraft glided to a safe
landing at Gimli, Manitoba \cite{acc:lockwood-gimli1985}.

\paragraph{Technical mechanism.}
The Boeing 767 was Air Canada's first metric-instrumented aircraft.
The fuel-quantity processor and the load-calculation workflow used
kilograms, while fuel itself was loaded by volume in litres.
Conversion required multiplication by the fuel's density. On the day
of the flight, the aircraft's fuel-quantity processor was
unserviceable, so the load was determined by an abnormal procedure
using drip sticks (mechanical fuel-volume gauges) plus a hand
calculation involving fuel density. The calculation was performed
using a factor near $1.77 \,\text{lb/L}$, the appropriate factor for
pounds-per-litre, where the kilogram-based calculation required
approximately $0.80 \,\si{\kilogram\per\liter}$. The result
overstated fuel mass by approximately a factor of $2.2$. The
aircraft departed with about half the fuel its flight plan required
\cite{acc:lockwood-gimli1985}. In flight, the fuel-monitoring system
reported low fuel and the engines flamed out from fuel exhaustion
before the aircraft reached Winnipeg. The crew used the 767's ram
air turbine for partial hydraulics and selected Gimli as a former
military airfield with a usable runway. The captain, an experienced
glider pilot, executed a forward slip and the aircraft landed
without engine power.

\paragraph{Organisational context.}
The Lockwood Board of Inquiry, established by Order in Council under
Justice George H.\ Lockwood, identified contributory factors beyond
the unit conversion. The fuel-quantity processor was unserviceable
in a manner that had been carried as a non-routine deferred defect.
Procedural shortcuts had been authorised under conditions that did
not match the day's actual maintenance state. Documentation
accompanying the metric-fuel transition assigned responsibility for
fuel-load verification across multiple roles in a way that allowed
each role to assume another role had verified the conversion. The
dispatch chain trusted the fuelling crew's arithmetic without an
independent cross-check \cite{acc:lockwood-gimli1985}. The
metric-fuel transition was the surrounding institutional event: a
Canadian airline was the first commercial 767 operator to be issued
metric-instrumented aircraft, and the unit conversion was a
predictable risk that no procedural barrier was tuned to catch.

\paragraph{Lesson for algebraic and proof discipline.}
The slip is a unit-arithmetic error of the kind a working engineer
recognises immediately: $\rho = 1.77$ has units of $\text{lb}/\si{
\liter}$, not $\si{\kilogram\per\liter}$, and the multiplication
$\text{volume} \times \rho$ produces mass in pounds, not in
kilograms. The crew's calculation never carried the unit through to
the answer. Had the form of the calculation been ``$V \times \rho =
m$, with $V$ in litres, $\rho$ in $\si{\kilogram\per\liter}$, $m$ in
kilograms,'' the missing $\si{\kilogram}/\text{lb}$ conversion would
have appeared as a dangling unit. The chapter's rule that every
quantity carries its unit through every step is the rule that, had
it been enforced in the fuelling worksheet, would have failed loudly
rather than silently. The case is the human-scale companion to the
Mars Climate Orbiter: the same notation failure, the same missing
audit at an interface, two orders of magnitude lower in
consequence.

\subsection{Ariane 5 Flight 501, 1996}

A third case shows that notation collapse extends beyond unit
mismatch into the precision and range of the data type itself. The
maiden flight of Ariane 5, on 4 June 1996, broke up about $39$
seconds after main engine ignition (about $31$ seconds after
lift-off) when an inertial reference system software exception,
triggered by a 64-bit-to-16-bit conversion overflow inside an
alignment function reused unchanged from Ariane 4, was unhandled by
both the primary and the back-up SRI; the launcher then commanded
extreme manoeuvres on bad attitude data and broke up under
aerodynamic loads \cite{acc:ariane501-ifr}.

\paragraph{Technical mechanism.}
The Inquiry Board chaired by Jacques-Louis Lions identified the
cause as an unhandled software exception in the inertial reference
system (SRI). The exception occurred during conversion of a 64-bit
floating-point representation of an internal horizontal-bias
variable (BH) to a 16-bit signed integer. The result exceeded the
integer's range $[-32768, 32767]$ and the conversion overflowed. The
exception was not caught; the SRI shut down, and the back-up SRI,
running the same software, shut down for the same reason. With both
SRIs offline, the on-board computer interpreted diagnostic output
from the SRI as flight data, computed extreme attitude corrections,
and the launcher commanded large nozzle deflections. Aerodynamic
loads exceeded structural limits and the vehicle broke up; the
self-destruct system then activated \cite{acc:ariane501-ifr}. The
conversion occurred inside an alignment function whose purpose was
to align the inertial reference platform on the launch pad before
lift-off. For Ariane 4, the function had been kept active for
several seconds after lift-off as part of a pre-launch-recovery
requirement. For Ariane 5, the same function had no operational
purpose after lift-off; the software was reused from Ariane 4
unchanged. Ariane 5 developed horizontal velocity at a rate that
took BH outside the 16-bit integer range; Ariane 4 had not. BH had
been one of seven variables that the Ariane 4 software protected
against overflow; the other three, including BH, had been left
unprotected on the basis that the Ariane 4 trajectory could not
produce out-of-range values. That analysis was not redone for Ariane
5.

\paragraph{Organisational context.}
The Inquiry Board found the failure to be a verification-domain
problem: software qualified for one launcher was reused on another
whose flight envelope had not been re-analysed against the reused
code. The board observed that the Ariane 5 qualification had treated
the reused SRI software as already qualified by Ariane 4 service
history, rather than as new software whose flight envelope must be
re-verified. The board recommended explicit reuse-with-modification
verification, including range analysis of all internal variables
against the new flight envelope, defensive programming for arithmetic
operations on flight-critical values, and software-architecture
review for reused components in qualification \cite{acc:ariane501-ifr}.
The relevant organisational pattern is reuse without re-verification:
a system component that performed correctly inside an old envelope
was lifted into a new envelope, and the burden of re-checking the
envelope was deferred on the assumption that prior service was
evidence enough.

\paragraph{Lesson for algebraic and proof discipline.}
Notation is more than units. A variable's representation includes
its precision (single, double, integer) and its range (the set of
values the type can hold). An algebraic identity that is valid in
$\R$ (the unbounded reals) is not automatically valid in the
finite-precision arithmetic of a fixed-width integer or a
fixed-precision float. The conversion of a 64-bit floating-point
horizontal-bias value to a 16-bit signed integer was an algebraic
operation that the writer had implicitly assumed to be a no-op,
because in the Ariane 4 flight domain the value was always in range.
The implicit hypothesis was not verified for the Ariane 5 flight
domain. The chapter's principle generalises to: every interface
variable carries domain, units, precision, range, and provenance,
and every reuse of a component across an envelope change re-verifies
each. The case is also the canonical reminder that the engineering
proof an algebra rule holds inside a real software system is a
range-and-precision proof, not only an identity proof.

\subsection{Two short-form cases on the same pattern}

The interface-as-place-where-notation-crosses pattern is recurrent
enough that two additional cases at short form are worth naming.

The Hubble Space Telescope's primary mirror, fabricated between 1979
and 1981 and delivered to orbit in April 1990, was figured to the
wrong shape because the reflective null corrector used to verify the
mirror during polishing was itself misassembled. A field lens inside
the null corrector was mispositioned along the optical axis by
approximately $1.3 \,\si{\milli\meter}$, because the spacing had
been set against the wrong feature of a metering rod: the end cap
and the target feature were not properly distinguished during
assembly. The polished mirror's outer edge was about $2 \,\si{\micro
\meter}$ too flat, an error large enough to ruin diffraction-limited
imaging across the optical bandpass. Auxiliary tests using a
refractive null corrector and an inverse null test had flagged
discrepancies during fabrication; both were set aside under the
assumption that the reflective null corrector was the more
authoritative instrument \cite{acc:nasa-hubble-optical-systems-1990}.
The Hubble case shares with Mars Climate Orbiter the structure of an
interface measurement that was not independently verified at the
boundary: in Hubble's case, the reference instrument was its own
unverified measurement chain, and the auxiliary readings that would
have caught the slip were dismissed.

The Patriot missile system at Dhahran, Saudi Arabia, on 25 February
1991, failed to track and intercept an incoming Scud missile because
of accumulated floating-point error in the system's tracking clock.
The clock variable was maintained as an integer count of tenths of
seconds and converted to a real number by multiplying by $0.1$,
which has no finite binary representation. After the battery had
been continuously powered for about $100$ hours, the accumulated
truncation error had drifted the time-of-flight calculation by
enough that the tracking gate fell outside the Scud's actual
location; the Scud passed through and impacted a barracks, killing
twenty-eight US Army soldiers \cite{acc:gao-patriot-1992}. The
notation slip here was the implicit assumption that the binary
representation of the decimal $0.1$ is exact, an assumption that
holds on paper and fails in fixed-precision binary arithmetic. The
mismatch between a decimal-natural interface specification and a
binary-floating-point implementation is the precision-and-range
analogue of the Mars Climate Orbiter unit slip. Both are notation
failures at an interface; both became visible only when the
cumulative error crossed an operational threshold.

\subsection{The pattern}

The three cases share a structure. A quantity passes from a
producing system to a consuming system. The quantity carries one
representation in the producing system and is interpreted in another
representation by the consuming system. The interface between the
two systems was the place where the representation should have been
audited. In each case, the audit was missing or weak. The
representation slipped between systems, and the slip was invisible
until it became consequential.

We state the discipline as the chapter's closing principle.

\begin{principle}[Notation as interface contract]
Every symbol in a calculation, every variable in a software
interface, every quantity in a handoff between systems carries an
explicit representation: domain, units, precision, range, and
provenance. The interface between two systems is the place where the
representation is audited, before the quantity is consumed. A
representation that is not audited at the interface is a
representation the consuming system has implicitly assumed.
\end{principle}

The Mars Climate Orbiter case is in the named-cases registry at
\repopath{docs/research/accidents/mars-climate-orbiter-1999.md}; the
Air Canada Flight 143 entry is at
\repopath{docs/research/accidents/air-canada-143-1983.md}; the
Ariane 5 entry is at
\repopath{docs/research/accidents/ariane-501-1996.md}; the Hubble
primary-mirror entry is at
\repopath{docs/research/accidents/hubble-primary-mirror-1990.md};
the Patriot Dhahran entry is at
\repopath{docs/research/accidents/patriot-dhahran-1991.md}. Each
entry documents the technical mechanism, the organisational
mechanism, and the primary or closest-equivalent source. The reader
will meet each of the five again in Volume X.

\section{Project}\pathtag{core}

\begin{project}[Rewrite a poorly-typeset engineering paper]
The reader chooses a real published engineering document of their
choice (a textbook chapter, a technical report, a journal article, a
standards document) whose notation has the kind of defects this
chapter has named. The reader rewrites one to three pages of the
document with the notation defects fixed.

\textbf{Track}. Analysis only. No physical artifact, no simulation,
no household component. \textbf{Hazard class}: none.

\textbf{Tools}. The chosen document, in a form the reader can mark
up. Paper, pencil, eraser. Optionally a typesetting environment
(LaTeX, Word, or another) to produce the rewrite.

\textbf{Procedure}. The reader selects a document and one to three
contiguous pages of it. Suggested defect classes to look for: a
symbol used before it is defined; a unit dropped from a quantity;
a quantity given in two unit systems without explicit conversion; an
index typeset ambiguously between subscript and exponent; an
overloaded symbol meaning two different things in the same passage;
an equation referenced without a number; a notation table absent
from a multi-symbol document. The reader marks every defect they
identify and produces a corrected rewrite of the page or pages.
Every symbol is introduced before use; every quantity carries its
unit; every variable appears in the rewrite's notation table; every
equation is numbered; every cross-reference uses the equation
number.

\textbf{Reflection (1500 words)}. After the rewrite, the reader
writes a 1500-word narrative addressing:

\begin{itemize}[leftmargin=*]
  \item Which defects were the most common in the chosen document.
  \item Which defects were the most consequential, in the sense
    that a reader could form an incorrect engineering interpretation
    from them.
  \item Which defects the reader's own writing also contains, and
    what habit of revision would catch them.
  \item Whether the document's audience (working engineers, students,
    a regulatory body, the general public) explains some of the
    defects, and whether the discipline of this chapter applies
    equally across audiences.
  \item One symbol or equation in the original document whose meaning
    the reader could not recover even after several readings, and
    what additional information would have made the meaning clear.
\end{itemize}

\textbf{Deliverable}. The original page or pages (photographed or
scanned, or referenced by citation), the marked-up version, the
rewrite, and the reflection. A reader who completes the project
will have produced an artifact that any later collaborator could
read and audit, and they will have noticed at least one notation
habit of their own that they had not noticed before.

\textbf{Time}. Two to three hours of selection and reading, three
to five hours of rewriting, one to two hours of reflection writing.
The total work fits in a long weekend.

\paragraph{Phases.}
The project unfolds in four phases.

\textit{Phase 1: selection (one to two hours).} The reader picks a
single published engineering document and selects one to three
contiguous pages whose notation has defects of the kind named in
section 1.6. A document with no defects is rare; a document with
many is usable, and the project's value rises with the density of
defects. Sources that often serve well: an older engineering paper
predating LaTeX conventions; a vendor whitepaper; a course handout;
a chapter of a translated textbook; an internal company technical
memo. A document with too many defects (a translated thesis, a
hand-written legacy report) overwhelms the rewrite; cap at one to
three pages.

\textit{Phase 2: defect inventory (one to two hours).} The reader
walks the page slowly and marks every defect by category: symbol
used before defined; unit dropped from a quantity; mixed unit
systems without explicit conversion; ambiguous index typesetting;
overloaded symbol; equation unnumbered; missing or absent notation
table. The reader keeps a tally per category; the tally is part of
the deliverable.

\textit{Phase 3: rewrite (three to five hours).} The reader produces
a corrected version of the same page or pages. Every symbol is
introduced before use; every quantity carries its unit; every
variable appears in the rewrite's notation table; every equation is
numbered; every cross-reference uses the equation number. The
rewrite preserves the original's content, not its prose; the reader
is rewriting the notation, not the argument.

\textit{Phase 4: reflection (one to two hours).} The 1500-word
narrative addresses the five questions stated above.

\paragraph{Deliverables.}
The project produces five artifacts:
\begin{enumerate}[leftmargin=*]
  \item The original page or pages, in a form a reviewer can
    consult (photograph, scan, or citation with page numbers).
  \item The marked-up version with every defect annotated.
  \item The defect inventory by category, with counts.
  \item The corrected rewrite.
  \item The 1500-word reflection.
\end{enumerate}

\paragraph{Success criteria.}
A successful project meets four criteria. The defect inventory
identifies at least eight defects across at least three categories.
The rewrite is readable by a reader who has not read the original;
the rewrite stands on its own. The reflection names at least one
notation habit of the reader's own writing that the project
surfaced. The reader can, six months later, return to the rewrite
and recover every symbol's meaning from the rewrite alone, without
recourse to the original.

\paragraph{Common failure modes.}
Three failure modes recur in this project. The reader chooses a
document that is already clean and produces a rewrite that
substantively repeats the original; the cure is to pick a document
that pre-dates the modern engineering-writing conventions.
The reader rewrites the argument rather than the notation,
``improving'' the technical content along the way; the cure is to
limit the rewrite to notation and to preserve every claim and every
derivation step exactly. The reader produces a rewrite cleaner than
their own habitual writing, then continues to write at the old
standard; the cure is the fourth question in the reflection, which
asks the reader to name a notation habit of their own and to
commit to a concrete change.
\end{project}

\section{Exercises}

The exercises mix calculation, derivation, estimation, simulation,
design, diagnosis, reverse engineering, failure analysis, and
judgement. Full
solutions are provided for the calculation and derivation exercises;
the estimation exercises receive published reference estimates with
the editor's working; the design, diagnosis, failure-analysis, and
judgement exercises receive rubrics rather than single answers.
Exercises are tagged with their reader-path tier: \emph{core} are
required, \emph{standard} are working-engineer default, and
\emph{mastery} use tools the chapter has not yet introduced and may
be deferred.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Expand $(a + b)^{4}$ using the binomial theorem and verify the
result by computing $(a+b)^{2} \cdot (a+b)^{2}$ as a separate
expansion. State which intermediate identity links the two.

\paragraph{Solution sketch.}
By the binomial theorem with $n = 4$, the coefficients are the
fourth row of Pascal's triangle, $1, 4, 6, 4, 1$, so
$(a+b)^{4} = a^{4} + 4 a^{3} b + 6 a^{2} b^{2} + 4 a b^{3} + b^{4}$.
By squaring twice, $(a+b)^{2} = a^{2} + 2ab + b^{2}$, then
$(a^{2} + 2ab + b^{2})^{2} = a^{4} + 4 a^{3} b + (2 + 4) a^{2} b^{2}
+ 4 a b^{3} + b^{4}$ after collecting the cross terms. The two
results match. The identity that links them is the binomial-
coefficient recurrence $\binom{4}{k} = \binom{2}{0}\binom{2}{k} +
\binom{2}{1}\binom{2}{k-1} + \binom{2}{2}\binom{2}{k-2}$, that is,
the Vandermonde identity for $\binom{2+2}{k}$.
\end{exercise}

\begin{exercise}
Factor each of the following over the real numbers: (a) $x^{4} - 16$,
(b) $x^{3} + 8$, (c) $x^{4} + 4$. For (c), use the Sophie Germain
identity $a^{4} + 4b^{4} = (a^{2} + 2ab + 2b^{2})(a^{2} - 2ab +
2b^{2})$. State for which the factorisation is over $\R$ and for
which it would require $\C$.

\paragraph{Solution sketch.}
(a) Difference of squares twice: $x^{4} - 16 = (x^{2} - 4)(x^{2} + 4)
= (x - 2)(x + 2)(x^{2} + 4)$. The remaining factor $x^{2} + 4$ has
no real roots; over $\C$ it factors as $(x - 2\mathrm{i})(x + 2\mathrm{i})$.
(b) Sum of cubes: $x^{3} + 8 = (x + 2)(x^{2} - 2x + 4)$. The
quadratic factor has discriminant $4 - 16 = -12 < 0$, so it is
irreducible over $\R$. (c) With $a = x$, $b = 1$ the Sophie Germain
identity gives $x^{4} + 4 = (x^{2} + 2x + 2)(x^{2} - 2x + 2)$, both
factors irreducible over $\R$ (their discriminants are $4 - 8 = -4
< 0$). All three factorisations live in $\R$ to the stated extent;
further factoring of (a)'s $x^{2} + 4$, (b)'s quadratic factor, and
(c)'s two quadratic factors requires $\C$.
\end{exercise}

\begin{exercise}
Compute the sum $\sum_{i=1}^{50} i^{2}$ using the closed form
$\sum_{i=1}^{n} i^{2} = \frac{n(n+1)(2n+1)}{6}$. State the value to
the exact integer.

\paragraph{Solution sketch.}
$\sum_{i=1}^{50} i^{2} = \frac{50 \cdot 51 \cdot 101}{6} =
\frac{257{,}550}{6} = 42{,}925$. The arithmetic is exact;
$50 \cdot 51 = 2550$, $2550 \cdot 101 = 257{,}550$, and
$257{,}550 / 6 = 42{,}925$. The formula is a closed form, so the
result is integer-exact without any rounding step.
\end{exercise}

\begin{exercise}
Solve the quadratic $2x^{2} - 7x + 3 = 0$ by factoring, by
completing the square, and by the quadratic formula. Confirm the
three approaches give the same two roots. State which approach you
would use first if the coefficients were $a = 1.7 \times 10^{-3}$,
$b = -2.4 \times 10^{2}$, $c = 1.1 \times 10^{4}$, and explain why.

\paragraph{Solution sketch.}
By factoring: $2 x^{2} - 7 x + 3 = (2x - 1)(x - 3)$, roots $x =
\tfrac{1}{2}, 3$. By completing the square: $x^{2} - \tfrac{7}{2} x
= -\tfrac{3}{2}$, $(x - \tfrac{7}{4})^{2} = -\tfrac{3}{2} +
\tfrac{49}{16} = \tfrac{25}{16}$, $x = \tfrac{7}{4} \pm \tfrac{5}{4}
= 3, \tfrac{1}{2}$. By the quadratic formula: $x = (7 \pm \sqrt{49 -
24})/4 = (7 \pm 5)/4 = 3, \tfrac{1}{2}$.

For the ill-conditioned case, the quadratic formula's
$\sqrt{b^{2} - 4ac}$ branch loses precision when $b^{2} \gg 4ac$
because the subtraction $-b \pm \sqrt{b^{2} - 4ac}$ produces
catastrophic cancellation in one of the two roots. With $b^{2} =
5.76 \times 10^{4}$ and $4ac = 4 \cdot 1.7 \times 10^{-3} \cdot 1.1
\times 10^{4} = 74.8$, the discriminant is $5.7525 \times 10^{4}$
and its square root is $\approx 239.85$; the root $x_{+} = (240 +
239.85)/(2 \cdot 1.7 \times 10^{-3}) \approx 1.41 \times 10^{5}$ is
fine but $x_{-} = (240 - 239.85)/(2 \cdot 1.7 \times 10^{-3})
\approx 44$ loses three significant figures in the subtraction. The
numerically safe approach is to compute the well-conditioned root
$x_{+}$ first and obtain the other from Vieta's relation $x_{-} =
c/(a x_{+}) = 1.1 \times 10^{4} / (1.7 \times 10^{-3} \cdot 1.41
\times 10^{5}) \approx 45.8$. The corrected calculation gives
$x_{-} \approx 45.8$, two digits different from the cancellation-
prone version. Vieta-first is the standard fix.
\end{exercise}

\begin{exercise}
A cantilever deflection formula is $\delta = \rho g L^{4} / (8 E
h^{2})$. With $\rho = 7850 \,\si{\kilogram\per\meter\cubed}$ (steel),
$g = 9.81 \,\si{\meter\per\second\squared}$, $L = 1.20 \,\si{\meter}$,
$E = 200 \,\si{\giga\pascal}$, $h = 8 \,\si{\milli\meter}$, compute
$\delta$ in millimetres. Carry units through every step and state
the final unit explicitly.

\paragraph{Solution sketch.}
Numerator: $\rho g L^{4} = 7850 \cdot 9.81 \cdot (1.20)^{4} =
77{,}009 \cdot 2.0736 \approx 1.597 \times 10^{5}$ in units of
$\si{\kilogram\per\meter\cubed} \cdot \si{\meter\per\second\squared}
\cdot \si{\meter\tothe{4}} = \si{\kilogram\meter\squared\per\second
\squared} = \si{\joule}$.
Denominator: $8 E h^{2} = 8 \cdot 200 \times 10^{9} \cdot (8 \times
10^{-3})^{2} = 8 \cdot 200 \times 10^{9} \cdot 6.4 \times 10^{-5} =
1.024 \times 10^{8}$ in units of $\si{\pascal} \cdot \si{\meter
\squared} = \si{\newton}$.
Ratio: $\delta = 1.597 \times 10^{5} / 1.024 \times 10^{8} = 1.56
\times 10^{-3} \,\si{\meter} = 1.56 \,\si{\milli\meter}$.
The unit balance $\si{\joule}/\si{\newton} = \si{\meter}$ is the
audit: the answer is a length, as expected for a deflection.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Prove the difference-of-squares identity $a^{2} - b^{2} = (a -
b)(a + b)$ from the distributive law alone. State at which step
commutativity is used.

\paragraph{Solution sketch.}
$(a - b)(a + b) = a(a + b) - b(a + b) = a^{2} + ab - ba - b^{2}$ by
two applications of distributivity. The cross terms cancel,
$ab - ba = 0$, because $a$ and $b$ commute; over a commutative ring
this step is the use of commutativity. The remaining terms give
$a^{2} - b^{2}$. Over a non-commutative ring, $ab \neq ba$ in
general and the identity fails: $(A - B)(A + B) = A^{2} + AB - BA -
B^{2}$, which equals $A^{2} - B^{2}$ only when $A$ and $B$ commute.
\end{exercise}

\begin{exercise}
Prove the two-variable AM--GM inequality $\frac{a + b}{2} \geq
\sqrt{ab}$ for $a, b \geq 0$ in two ways: (a) by squaring both
sides, (b) by writing $a = (\sqrt{a})^{2}$ and $b = (\sqrt{b})^{2}$
and using $(\sqrt{a} - \sqrt{b})^{2} \geq 0$. State which proof
generalises more naturally to the $n$-variable case.

\paragraph{Solution sketch.}
(a) Both sides are non-negative, so squaring preserves the
inequality: $(a + b)^{2}/4 \geq ab$ iff $(a + b)^{2} \geq 4 ab$ iff
$a^{2} - 2 ab + b^{2} = (a - b)^{2} \geq 0$, which is always true;
equality when $a = b$.
(b) Write $u = \sqrt{a}$, $v = \sqrt{b}$ (both non-negative). Then
$(u - v)^{2} \geq 0$ expands to $u^{2} - 2 u v + v^{2} \geq 0$, that
is, $a - 2 \sqrt{ab} + b \geq 0$, giving $\tfrac{a + b}{2} \geq
\sqrt{ab}$; equality when $u = v$.
Proof (a), seen as $(a + b)^{2} \geq 4 ab$, generalises to $n$
variables only with extra structure (Cauchy-style polarisation).
Proof (b), seen as the convexity of the function $x \mapsto -\log
x$, generalises directly to $n$ variables by Jensen's inequality:
$\log(\tfrac{1}{n}\sum a_{i}) \geq \tfrac{1}{n} \sum \log a_{i}$,
which is the $n$-variable AM-GM after exponentiation. The
log-convexity proof is the natural generalisation.
\end{exercise}

\begin{exercise}
Derive the Cauchy--Schwarz inequality from the AM--GM inequality
applied term by term. (Hint: bound each $a_{i} b_{i}$ by
$\tfrac{1}{2}(a_{i}^{2}/\lambda + \lambda b_{i}^{2})$ for a free
parameter $\lambda > 0$, then sum and minimise over $\lambda$.)

\paragraph{Solution sketch.}
For each $i$, AM-GM applied to $a_{i}^{2}/\lambda$ and $\lambda
b_{i}^{2}$ (both non-negative) gives $\sqrt{a_{i}^{2}/\lambda \cdot
\lambda b_{i}^{2}} = \abs{a_{i} b_{i}} \leq \tfrac{1}{2}(a_{i}^{2}/
\lambda + \lambda b_{i}^{2})$. Summing over $i$,
\[
\sum_{i} \abs{a_{i} b_{i}} \leq \frac{1}{2 \lambda} \sum_{i}
a_{i}^{2} + \frac{\lambda}{2} \sum_{i} b_{i}^{2}.
\]
The right side is minimised in $\lambda$ at $\lambda^{2} = (\sum
a_{i}^{2}) / (\sum b_{i}^{2})$ (set derivative to zero), giving the
minimum value $\sqrt{(\sum a_{i}^{2})(\sum b_{i}^{2})}$. Since
$\abs{\sum a_{i} b_{i}} \leq \sum \abs{a_{i} b_{i}}$ by the triangle
inequality, $\abs{\sum a_{i} b_{i}} \leq \sqrt{(\sum a_{i}^{2})(\sum
b_{i}^{2})}$, which on squaring is Cauchy-Schwarz. The trick is the
classical AM-GM-with-free-parameter argument; the parameter
$\lambda$ rescales the two halves until they balance.
\end{exercise}

\begin{exercise}
Prove by induction that $\sum_{i=1}^{n} (2i - 1) = n^{2}$, that is,
that the sum of the first $n$ odd integers is $n^{2}$. State the
base case and the inductive step explicitly.

\paragraph{Solution sketch.}
Base case $n = 1$: left side $= 2 \cdot 1 - 1 = 1$; right side $=
1^{2} = 1$.
Inductive step: assume $\sum_{i=1}^{n} (2i - 1) = n^{2}$. Add the
$(n+1)$-th term, $2(n+1) - 1 = 2n + 1$, to both sides:
$\sum_{i=1}^{n+1} (2i - 1) = n^{2} + 2n + 1 = (n+1)^{2}$.
The identity holds for $n+1$, and by induction for every $n \geq 1$.
The geometric picture: the sum of the first $n$ odd integers builds
the $n$-th square by adding an L-shaped gnomon of width $2n + 1$ at
each step.
\end{exercise}

\begin{exercise}
Prove by induction that $\binom{n+1}{k} = \binom{n}{k-1} +
\binom{n}{k}$ for $1 \leq k \leq n$. Use a direct algebraic
manipulation rather than a combinatorial argument, and state where
the algebraic identity $a + b = \frac{a(k) + b(n+1-k)}{n+1} \cdot
(n+1) / k$ enters.

\paragraph{Solution sketch.}
The identity is in fact direct, no induction required, but we
write the algebra step for $1 \leq k \leq n$:
\[
\binom{n}{k-1} + \binom{n}{k}
= \frac{n!}{(k-1)!(n-k+1)!} + \frac{n!}{k!(n-k)!}.
\]
Common denominator $k!(n-k+1)!$ gives
\[
= \frac{n! \cdot k}{k!(n-k+1)!} + \frac{n! \cdot (n-k+1)}{k!(n-k+1)!}
= \frac{n! \, [k + (n-k+1)]}{k!(n-k+1)!}
= \frac{n!\,(n+1)}{k!(n-k+1)!}
= \frac{(n+1)!}{k!(n+1-k)!}
= \binom{n+1}{k}.
\]
The hint identity captures the weighting step: the two terms are
combined with weights $k$ and $n - k + 1$ to clear the denominators.
Induction is needed only if one wants a proof of the general
binomial theorem from this recurrence; the recurrence itself is
algebraic.
\end{exercise}

\begin{exercise}
Prove the reverse triangle inequality $\abs{\,\abs{a} - \abs{b}\,}
\leq \abs{a - b}$ from the triangle inequality. (Hint: write $a = (a
- b) + b$ and apply the triangle inequality to $\abs{a}$, then
repeat with $a$ and $b$ swapped.)

\paragraph{Solution sketch.}
$\abs{a} = \abs{(a - b) + b} \leq \abs{a - b} + \abs{b}$ by the
triangle inequality, so $\abs{a} - \abs{b} \leq \abs{a - b}$.
Swapping $a$ and $b$: $\abs{b} - \abs{a} \leq \abs{b - a} = \abs{a - b}$.
The two together give $\abs{\,\abs{a} - \abs{b}\,} \leq \abs{a - b}$
because $\abs{x}$ is the larger of $x$ and $-x$. Equality holds when
$a$ and $b$ have the same sign and one of them is zero, the same
condition as the forward triangle inequality.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the order of magnitude of $50!$ (fifty factorial) without
direct computation. State your method, give a factor-of-three
confidence range, then compute the exact value (or look it up) and
report the ratio of your estimate to the true value.

\paragraph{Solution sketch.}
Stirling: $n! \approx \sqrt{2 \pi n}\,(n/e)^{n}$. For $n = 50$,
$50/e \approx 50/2.718 \approx 18.4$. So $50! \approx \sqrt{100 \pi}
\cdot 18.4^{50}$. $\sqrt{100 \pi} \approx 17.7$. $\log_{10}(18.4)
\approx 1.265$, so $18.4^{50} \approx 10^{63.2}$, and the prefactor
contributes $\log_{10}(17.7) \approx 1.25$. Estimate: $50! \approx
10^{64.5} \approx 3 \times 10^{64}$; factor-of-three range $10^{64}$
to $10^{65}$. The exact value is $50! = 3.04 \times 10^{64}$. The
ratio of estimate to truth is about $1.0$; Stirling is accurate at
$n = 50$ to better than $0.2\%$, because its relative error scales
as $1/(12 n) \approx 0.17 \%$.
\end{exercise}

\begin{exercise}
A fair coin is tossed $n = 100$ times. Estimate the probability that
the number of heads lies in the range $[40, 60]$, using a normal
approximation reasoning at order-of-magnitude level. State the
mean, the standard deviation, and the working approximation.

\paragraph{Solution sketch.}
For a binomial $\text{Bin}(n, p)$ with $n = 100$, $p = 1/2$, the
mean is $\mu = np = 50$ and the variance is $np(1-p) = 25$, so the
standard deviation is $\sigma = 5$. The range $[40, 60]$ is $\mu
\pm 2 \sigma$, the canonical two-sigma band of the normal
approximation. Under the normal approximation, the probability of
falling within $\pm 2 \sigma$ is approximately $0.954$; with a
continuity correction the binomial answer is closer to $0.965$. The
working approximation is: ``two-sigma for a binomial $\approx 95
\%$.'' The estimate is correct to about $1 \%$ in absolute
probability.
\end{exercise}

\begin{exercise}
The compound interest formula $(1 + r)^{n}$ is the engineer's
shortest exposure to growth at rate $r$ over $n$ periods. For $r =
0.05$ and $n = 30$, estimate $(1.05)^{30}$ by Bernoulli's
inequality (lower bound), by the linear approximation $1 + nr$
(both bounds), and by recognising $(1 + r)^{n} \approx
e^{nr}$ for small $r$. Compare the three estimates to a calculator
value.

\paragraph{Solution sketch.}
Bernoulli (lower bound): $(1.05)^{30} \geq 1 + 30 \cdot 0.05 = 2.50$.
Linear: $1 + nr = 2.50$ as a typical-value approximation; the
quadratic correction $\binom{n}{2} r^{2} = 435 \cdot 0.0025 = 1.09$
brings it to about $3.59$.
Exponential: $(1.05)^{30} \approx e^{30 \cdot 0.05} = e^{1.5}
\approx 4.48$.
Exact: $(1.05)^{30} \approx 4.32$. The Bernoulli bound is loose by
about $42 \%$. The two-term Taylor expansion is closer at $16 \%$
under. The exponential approximation overshoots by about $4 \%$;
$e^{nr}$ is the limit of $(1 + r)^{n}$ as $r \to 0$ with $nr$ fixed,
and at $r = 0.05$ the difference is the quadratic-and-higher
correction. The exponential is the most accurate of the three for
typical engineering rates over moderate time horizons.
\end{exercise}

\begin{exercise}
Estimate the number of distinct strings of length $20$ over an
alphabet of $4$ letters. Express the result in the closest power of
$10$. Then estimate the number of length-$20$ strings in which
exactly $10$ letters are vowels (here, two specific letters are
vowels and two are consonants). State the binomial coefficient that
governs the count and approximate it.

\paragraph{Solution sketch.}
Total strings: $4^{20} = (2^{2})^{20} = 2^{40} \approx 10^{12}$
(more precisely, $1.10 \times 10^{12}$). Strings with exactly $10$
vowels and $10$ consonants: choose which positions are vowels
($\binom{20}{10} = 184{,}756$), assign one of two vowels at each
($2^{10}$), assign one of two consonants at each ($2^{10}$). Total
$= \binom{20}{10} \cdot 2^{10} \cdot 2^{10} = 184{,}756 \cdot 2^{20}
\approx 1.85 \times 10^{5} \cdot 1.05 \times 10^{6} \approx 1.94
\times 10^{11}$. The binomial $\binom{20}{10}$ is the central
binomial coefficient; it grows as $4^{n}/\sqrt{\pi n}$ for large
$n$, giving the approximation $\binom{20}{10} \approx 4^{10}/
\sqrt{10 \pi} \approx 10^{6}/5.6 \approx 1.86 \times 10^{5}$ (the
Stirling answer), within $1 \%$ of the exact value.
\end{exercise}

\begin{exercise}
A working engineer is asked, at a meeting, for the order of magnitude
of $\sqrt{n!}$ where $n = 100$. Estimate the answer in the closest
power of $10$, using Stirling's approximation $n! \approx \sqrt{2\pi
n}\,(n/e)^{n}$ as the working tool. State the dominant term and the
error class in your estimate.

\paragraph{Solution sketch.}
$n! \approx \sqrt{2 \pi n}(n/e)^{n}$, so $\sqrt{n!} \approx (2 \pi
n)^{1/4} (n/e)^{n/2}$. For $n = 100$: $n/e \approx 36.8$,
$(n/e)^{n/2} = 36.8^{50}$; $\log_{10}(36.8) \approx 1.566$, so
$36.8^{50} \approx 10^{78.3}$. The prefactor $(200 \pi)^{1/4}
\approx 5.0$, contributing $\log_{10}(5) \approx 0.7$. So $\sqrt{100!}
\approx 10^{79}$. The dominant term is $(n/e)^{n/2}$ at $n = 100$;
the prefactor contributes less than one decimal digit. The error
class is Stirling's, about $1/(12 n) \approx 0.08 \%$ in $n!$,
half of that in $\sqrt{n!}$, which is irrelevant at the order-of-
magnitude level the question asked.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
In a programming environment of the reader's choice, write a
function that takes integers $n$ and $k$ and returns $\binom{n}{k}$
using only integer arithmetic (no floating point). Use the recursion
$\binom{n}{k} = \binom{n-1}{k-1} + \binom{n-1}{k}$ with memoisation,
and verify your implementation against $\binom{52}{5} = 2{,}598{,}960$
(the number of five-card poker hands) and $\binom{30}{15} =
155{,}117{,}520$.

\paragraph{Solution sketch.}
The reference implementation lives at
\texttt{code/binomial\_table.py}. The core function uses a
dictionary $\{(n, k) \to \binom{n}{k}\}$ and the Pascal recurrence,
with boundary conditions $\binom{n}{0} = \binom{n}{n} = 1$ and
$\binom{n}{k} = 0$ for $k < 0$ or $k > n$. Running with
\texttt{max\_n = 52} writes a CSV from which $\binom{52}{5}$ and
$\binom{30}{15}$ can be looked up; both match the stated values
exactly. The integer-only implementation avoids the catastrophic
precision loss that the closed form $n!/(k!(n-k)!)$ suffers when
$n!$ overflows the machine word, and avoids the floating-point
rounding that the gamma-function-based implementation introduces.
\end{exercise}

\begin{exercise}
Verify Cauchy--Schwarz numerically. Generate $1000$ pairs of vectors
$u, v \in \R^{10}$ from a standard normal distribution. For each
pair, compute $\inner{u, v}^{2}$ and $\norm{u}^{2} \norm{v}^{2}$,
and confirm the first is at most the second. Plot the ratio of the
two as a histogram; report the maximum ratio observed and explain
why it should be close to $1$ but not exceed it.

\paragraph{Solution sketch.}
The reference implementation lives at
\texttt{code/cauchy\_schwarz\_mc.py}. With $d = 10$ and $N = 1000$
under the fixed random seed, the maximum observed ratio is
approximately $0.65$ and the mean is approximately $1/d = 0.10$,
matching the prediction that the squared cosine of the angle between
two independent isotropic Gaussian vectors in $\R^{d}$ concentrates
around $1/d$. No trial exceeds $1$; the bound is a theorem, and the
script's assertion catches any implementation bug that would let it.
The maximum stays well below $1$ at large $d$ because two random
isotropic vectors are nearly orthogonal with high probability in
high dimensions, the standard concentration of measure picture.
\end{exercise}

\begin{exercise}
The Collatz iteration takes a positive integer $n$ to $n/2$ if $n$
is even and to $3n + 1$ if $n$ is odd, repeatedly. The Collatz
conjecture states that the iteration reaches $1$ from every starting
$n$. Implement the iteration and run it for every starting value
$n = 1, 2, \ldots, 10^{6}$. Report the maximum number of steps to
reach $1$ and the starting value that achieves it. State explicitly
that the simulation is not a proof of the conjecture and explain
why.

\paragraph{Solution sketch.}
The reference implementation lives at \texttt{code/collatz\_run.py}.
For $n$ up to $10^{6}$ the maximum step count is $524$, achieved at
$n = 837{,}799$. The simulation is not a proof: the Collatz
conjecture (current as of 2026) is an open problem in mathematics.
A run over the first $10^{6}$ starting values certifies the
conjecture on that finite set, not on the integers. The smallest
unresolved starting value, were one to exist outside the simulated
range, could be the next integer above $10^{6}$. The chapter's
distinction between demonstration and proof is the point of the
exercise; the demonstration is useful for code testing, for building
intuition about the orbit structure, and for catching an
implementation bug if a step count overflows. It is not a result
about the integers.
\end{exercise}

\begin{exercise}
Implement the AM--GM inequality check for vectors of length $n$
drawn from a non-negative distribution of the reader's choice (for
example, the absolute value of a standard normal, or the exponential
distribution with rate $1$). For $n \in \{2, 5, 10, 50\}$, generate
$10^{4}$ samples and report the empirical fraction by which the
arithmetic mean exceeds the geometric mean. Plot the ratio
$(\text{AM})/(\text{GM})$ as a function of $n$ and comment on its
behaviour.

\paragraph{Solution sketch.}
The reference implementation lives at \texttt{code/am\_gm\_check.py}.
For the exponential distribution with rate $1$, the script reports
empirical means of the ratio AM/GM that increase with $n$: about
$1.40$ at $n = 2$, $1.81$ at $n = 5$, $2.13$ at $n = 10$, and
about $4.3$ at $n = 50$. The ratio grows roughly like $\sqrt{n}$ at
moderate $n$ for the exponential distribution, and the variance of
the ratio falls as $n$ grows. Every sample respects the inequality:
the fraction with AM $\geq$ GM is $1.0$ (every sample is non-strict
in the equality case where all $n$ values happen to be equal, which
has probability zero in continuous distributions). AM-GM is a
theorem; the simulation cannot violate it. The growth of the ratio
with $n$ reflects the spread of the distribution: with more samples,
the chance of a small value pulling the geometric mean down rises.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a notation table for a hypothetical engineering paper on heat
conduction in a slab. The paper uses, at minimum, mass density,
thermal conductivity, specific heat capacity, temperature, time,
spatial coordinate, heat flux, and the thermal diffusivity that
combines several of the previous. Specify each symbol, its meaning,
its domain or type, its units, and the section of the paper where
it is first introduced. State which two symbols are most at risk of
being overloaded and what convention prevents the overload.

\paragraph{Discussion.}
A reasonable working table assigns: $\rho$ for mass density
($\si{\kilogram\per\meter\cubed}$), $k$ for thermal conductivity
($\si{\watt\per\meter\per\kelvin}$), $c_{p}$ for specific heat at
constant pressure ($\si{\joule\per\kilogram\per\kelvin}$), $T$ for
temperature ($\si{\kelvin}$ or $\si{\degreeCelsius}$ with stated
convention), $t$ for time ($\si{\second}$), $x$ for spatial
coordinate ($\si{\meter}$), $q$ for heat flux ($\si{\watt\per\meter
\squared}$), $\alpha = k / (\rho c_{p})$ for thermal diffusivity
($\si{\meter\squared\per\second}$). Two overloading risks dominate:
$k$ is also a common index variable; the convention is to reserve
$k$ for the conductivity in this paper and use $i$ or $j$ for
indices. $T$ is also the natural symbol for the time period of a
periodic boundary condition; the convention is to reserve $T$ for
temperature and use $\tau$ for the time period.
\end{exercise}

\begin{exercise}
Design a one-page algebraic identity reference card for a working
engineer. The card lists at most ten identities and states, for
each, the conditions under which it holds and the most common
application. Defend your choice of ten identities against an
imagined competitor card.

\paragraph{Discussion.}
The chapter's working library of six is the obvious starting set:
distributivity, square of a sum, difference of squares, cube of a
sum, geometric finite sum, telescoping. Four further candidates earn
their place by frequency in working practice: the binomial theorem
for general $n$; the small-$x$ expansion $(1 + x)^{n} \approx 1 +
nx + \tfrac{n(n-1)}{2} x^{2}$; the geometric infinite sum
$1/(1 - r)$ for $\abs{r} < 1$; and either Vieta's relations or the
quadratic formula with its Vieta companion $x_{1} x_{2} = c/a$.
The defence against a competitor card that includes, say, the
double-angle formulas or partial-fraction decomposition is that the
card is for the algebraic identities that recur outside trigonometry
and calculus; trigonometric identities live on a separate card
(Chapter 3 of this volume), and partial fractions live with calculus
(Chapter 6). The discipline of the card is to keep its scope
explicit and to refuse to grow it past one page.
\end{exercise}

\begin{exercise}
Design a checklist that an engineer reviewing a junior colleague's
calculation would walk down to verify the algebra. The checklist
has at most twelve items, each of which is something the reviewer
checks and the writer should have already checked. State which
three items are most often missed in your experience or in the
chapter's discussion.

\paragraph{Discussion.}
A defensible twelve-item checklist: (1) every symbol defined before
use; (2) every quantity carries its unit at every step; (3) the
final unit matches the expected dimension; (4) sign conventions
stated and consistent; (5) cancellation moves cite the non-zero
condition; (6) inequality directions audited at every multiplicative
step; (7) approximations bound their error; (8) order-of-magnitude
estimate precedes the calculation and is within a factor of three of
the answer; (9) every equation is numbered if referenced; (10) every
cross-reference uses the number; (11) intermediate factors
(e.g.\ the cross term $2ab$ in a square of a sum) checked; (12)
exponent rules audited for $e^{x}$ and other transcendental cases.
The three most often missed, by the chapter's evidence: (2) units at
every step (the Mars Climate Orbiter and Air Canada cases); (8) the
estimation before the calculation (the failure to catch a
$2 \times$ slip without an independent estimate); and (11)
intermediate cross terms (the $V_{0}^{2}$ exercise of section 1.1).
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{standard}

\begin{exercise}
A colleague's derivation arrives at a result that disagrees with the
expected answer by a factor of $2$. Without knowing which step
contains the error, propose a sequence of three checks that an
auditor could run to localise the error. (Hint: dimensional analysis
first, then recompute one summand, then compare against an
independent estimation.)

\paragraph{Solution sketch.}
Step 1: dimensional analysis. Verify both sides of the result carry
the same units. A factor-of-2 slip cannot arise from a unit error
(those produce orders of magnitude), so this is a fast eliminate-
the-easy-explanation pass. Step 2: recompute the most numerically
involved single substitution. A common factor-of-2 slip is dropping
the $\tfrac{1}{2}$ in $\tfrac{1}{2} m v^{2}$, $\tfrac{1}{2} C V^{2}$,
$\tfrac{1}{2} L I^{2}$, the cross term $2 a b$ in $(a + b)^{2}$, or
the integral $\int x \,\dd x = \tfrac{1}{2} x^{2}$. Step 3:
independent estimation. Estimate the answer's order of magnitude
without using the same derivation; if the colleague's result is
$2 \times$ the estimate and the estimate is internally consistent,
the slip is in the derivation. If the estimate matches the
colleague's result, the slip is in the expected value.
\end{exercise}

\begin{exercise}
A formula in a textbook reads $E = \tfrac{1}{2} m v^{2}$ in one
chapter and $E = m v^{2}$ in another, with no comment. Diagnose
the most likely explanation (different definition of $E$, missing
factor, typo, different physical regime) and state what evidence
you would look for in the surrounding text to settle the question.

\paragraph{Solution sketch.}
The most likely explanation is a different physical regime or a
different definition of $E$. The factor-of-2 is the canonical
non-relativistic-versus-relativistic-or-time-averaged distinction.
Specifically: in relativistic mechanics, the total energy of a
particle is $E = \gamma m c^{2}$, not $\tfrac{1}{2} m v^{2}$; the
$E = m v^{2}$ formula might be a time-averaged kinetic energy in a
context where $v$ is a peak velocity and the factor of $\tfrac{1}{2}$
disappears (it does for a sinusoidal velocity: $\langle v^{2} \rangle
= \tfrac{1}{2} v_{\text{peak}}^{2}$, and the kinetic energy of the
peak motion is $\tfrac{1}{2} m v_{\text{peak}}^{2}$, so the time-
averaged kinetic energy is $\tfrac{1}{4} m v_{\text{peak}}^{2}$;
the $E = m v^{2}$ form is not this). The likeliest scenario is a
typesetting error (a dropped $\tfrac{1}{2}$) or a missing definition
of $v$ as an RMS rather than a peak. Evidence to settle: the
context's stated definition of $v$; the worked example's numerical
agreement with one form; the dimensional check ($E$ in joules
either way); a sanity check against a known case.
\end{exercise}

\begin{exercise}
A spreadsheet computes the mass of a cylindrical pin from $m = \rho
\pi r^{2} L$. The output is a factor of $10^{3}$ too large. State
the three most likely places the error sits (units of $\rho$, units
of $r$, units of $L$) and the diagnostic check that distinguishes
them.

\paragraph{Solution sketch.}
A factor of $10^{3}$ is one of three slips. (1) $\rho$ entered in
$\si{\gram\per\centi\meter\cubed}$ rather than $\si{\kilogram\per
\meter\cubed}$ ($\rho_{\text{Fe}} = 7.85$ vs $7850$). The
spreadsheet then overcomputes by exactly $10^{3}$. (2) $r$ entered
in $\si{\milli\meter}$ where the formula expects $\si{\meter}$;
since $r$ enters squared, the resulting error is $10^{6}$, not
$10^{3}$. Wrong factor. (3) $L$ entered in $\si{\milli\meter}$
where the formula expects $\si{\meter}$; enters linearly, gives
$10^{3}$. The discriminating check: divide the output mass by the
geometric volume computed in clearly stated SI units. If the ratio
is exactly $7.85$, the density slip is the cause; if exactly $7850$,
the geometry slip is the cause. The diagnostic separates a unit-of-
$\rho$ error from a unit-of-$L$ error in one calculation.
\end{exercise}

\begin{exercise}
A junior engineer presents an inductive proof that all real numbers
$x$ satisfy $x = 0$. The base case is $0 = 0$. The inductive step
takes ``$x = 0$ implies $x + \epsilon = 0$ for every $\epsilon$.''
The conclusion: every real number equals zero. Locate the logical
defect with precision. State which property of induction the proof
violates and why ordinary induction does not extend to the real
numbers.

\paragraph{Solution sketch.}
The inductive step is simply false: $x = 0$ does not imply $x +
\epsilon = 0$; it implies $x + \epsilon = \epsilon$. The proof
arrives at a true base case followed by a false inductive step.
Even were the inductive step true, the proof would fail
structurally: ordinary induction extends a statement from $n$ to
$n + 1$ along the natural numbers, which are discrete and well-
ordered. The real numbers are dense and have no successor relation;
no inductive step of the form ``$P(x) \Rightarrow P(x + \epsilon)$''
can cover every real number, because the iterates $0, \epsilon, 2
\epsilon, \ldots$ form a discrete subset of $\R$ that misses every
irrational multiple of $\epsilon$. The defect is the implicit
assumption that induction works on a continuum. It does not;
continuum statements are established by analytic methods (continuity,
limits, supremum arguments) rather than by induction.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Read the executive summary of the Mars Climate Orbiter mishap report
\cite{acc:nasa-mco-mib}. State (a) the named root cause, (b) at
least two contributing factors the report identifies but does not
call root, and (c) one notation discipline from this chapter that,
applied at the contractor-to-operator interface, would have caught
the unit slip.

\paragraph{Solution sketch.}
(a) The named root cause is the failure to use metric units in the
coding of the ground software file \texttt{SM\_FORCES} for small-
forces data. (b) Contributing factors named in the report include:
inadequate consideration of the entire mission system in trajectory
correction manoeuvre design; inadequate operational navigation team
interaction with the spacecraft operations team; inadequate process
for verification, validation, and certification of trajectory data;
inadequate small-forces data analysis during cruise. (c) The
notation discipline of section 1.6 that would have caught the slip:
``units in every quantity, audited at the interface.'' Either an
interface-table audit (as in
\texttt{data/mco-unit-handoff.csv}) or a unit-checked transport of
the impulse value would have flagged the dimensional mismatch
before the first manoeuvre.
\end{exercise}

\begin{exercise}
The Air Canada Flight 143 incident involved a unit conversion error
inside a fuelling calculation \cite{acc:lockwood-gimli1985}. State
which of the notation rules from section 1.6 the fuelling
calculation violated. Argue, in 200 to 300 words, whether the
notation discipline alone would have prevented the incident, or
whether procedural and instrument-availability changes were also
needed.

\paragraph{Solution sketch.}
The most directly violated rule from section 1.6 is ``units in every
quantity''. The crew's calculation form was $V \times \rho = m$
without the unit chain attached, so $V$ in litres multiplied by
$\rho$ in $\text{lb}/\text{L}$ produced $m$ in pounds, which the
crew then read as kilograms. A unit-bearing form, $V \,[\text{L}]
\times \rho \,[\text{kg}/\text{L}] = m \,[\text{kg}]$, would have
required the conversion factor $\text{kg}/\text{lb} = 0.4536$ to
appear explicitly as a multiplier, surfacing the dimensional
mismatch.

The 200-300 word argument: notation discipline alone is necessary
but not sufficient. The fuelling crew worked under three
simultaneous defects: the fuel-quantity processor was unserviceable;
the abnormal procedure for drip-stick-based load calculation had
been authorised without an independent unit-cross-check; the
metric-fuel transition had assigned conversion responsibility across
roles in a way that allowed each role to assume another verified.
A unit-bearing calculation form would have caught the slip at the
calculation step. But the broader institutional context, in which a
deferred-defect entry was carried without escalating to a hold,
contained several latent failures that the unit-discipline fix
would not have addressed. The Lockwood report's recommendations
included instrument-redundancy improvements, procedural revisions
for abnormal fuel loads, and explicit cross-verification roles
during the metric-fuel transition. Notation discipline is one of
several barriers; the case shows that systems-level defences
include but are not limited to algebraic hygiene.
\end{exercise}

\begin{exercise}
The Ariane 5 Flight 501 inquiry report names the cause as a software
exception in a routine that converted a 64-bit floating-point number
to a 16-bit signed integer \cite{acc:ariane501-ifr}. Argue, in 200
to 300 words, that this is a notation-discipline failure of the
kind named in section 1.6, and identify which rule (definition
before use, units, indices, overloading, equation numbering,
notation table) the case most directly violates.

\paragraph{Solution sketch.}
The Ariane 5 case extends notation discipline from units to the
representation (precision and range) of a variable. The variable
BH (horizontal bias) was defined as a 64-bit floating-point value in
the alignment function inherited from Ariane 4, and was implicitly
expected to fit a 16-bit signed integer at the conversion site.
The implicit hypothesis ``the value is always in the integer
range'' was true for Ariane 4 and false for Ariane 5; the
hypothesis was not stated as a precondition on the conversion, was
not re-verified for the new launcher, and was not protected by a
range check.

Read against section 1.6: the rule most directly violated is
``definition before use,'' extended to mean ``definition of the
representation, including domain, range, and precision, audited at
every reuse.'' A notation table that listed BH with its expected
range, and a discipline of re-deriving the range bound under the
new flight envelope, would have surfaced the overflow risk before
launch. The chapter's principle on interface contracts reads the
case as the canonical reminder that an implicit precondition is
not a contract; a contract is what the producing and consuming
code can both verify. The case adds to notation discipline a rule
the chapter prose makes explicit: reuse across envelope changes
re-verifies every internal variable's range and precision against
the new envelope. The 220-word version of the argument carries the
same content with the empirical context (Ariane 4 service history,
Lions Inquiry Board) cited from the report.
\end{exercise}

\subsection*{Judgement}\pathtag{standard}

\begin{exercise}
A working engineer says: ``Proofs are for mathematicians; we just
need formulas that work.'' Write a one-paragraph reply that takes
the colleague seriously, agrees where agreement is honest, and
disagrees where disagreement is honest. State at least one specific
class of engineering problem in which a proof is the only defensible
answer.

\paragraph{Discussion.}
A defensible reply: the colleague is right about most engineering
work, which is calculation, derivation, and measurement against
formulas the discipline has already validated. The reply gives
ground there. Where the colleague is wrong is in cases where the
property the engineer is asserting cannot be observed only by
running the system: a numerical scheme's stability for every initial
condition in a class; a cryptographic protocol's resistance to every
attack from a stated class; a fault-tolerant algorithm's
correctness under every interleaving of failures; a safety
property of a piece of medical-device firmware. In each of these,
testing covers a finite sample of an infinite space, and the
remaining space is where the system will eventually be operated.
The proof is what closes the gap between tested behaviour and
deployed behaviour. The chapter's reading: proofs are tools for
load-bearing claims that experiment cannot reach, and the engineer
who can tell which claims need proofs and which do not has the
working judgement the discipline asks for.
\end{exercise}

\begin{exercise}
Read a single page of a textbook in your own engineering domain
(structural, electrical, software, chemical, biomedical, civil,
materials). Identify, in writing, three notation choices the
textbook makes and assess each against the rules of section 1.6.
State which choice you would change and why.

\paragraph{Discussion.}
The exercise is a working application of section 1.6, with no
single answer; the reader's submission stands or falls on how
specifically the chosen notation choices are named and how
specifically the assessment uses the chapter's rules. A strong
submission names three concrete choices (for instance: the textbook
uses the same symbol $T$ for temperature and for tension; the
textbook does not include a notation table; the textbook indexes
matrix entries as $a_{ij}$ without distinguishing row-then-column
from column-then-row convention) and assesses each against the
relevant rule (overloading, notation table, indices typeset
properly). A weak submission stays at the level of ``the symbols
are clear,'' which is not an assessment. The reader who finds a
textbook with no flaws is asked to assess a different textbook;
the discipline of the exercise is the assessment, not the verdict.
\end{exercise}

\begin{exercise}
Argue, in 200 to 300 words, whether the discipline of writing every
calculation with units carried through every step is too much
overhead for a working engineer to maintain. Take a position. Defend
it.

\paragraph{Discussion.}
The chapter's position: not too much overhead, given how cheaply the
discipline catches the kind of error that compounds invisibly. The
defensible counterargument is that carrying units in every
back-of-envelope estimation is overkill; a working engineer's
fluency includes the ability to drop units when the calculation is
short and confined to a single unit system, and to add them back
when the calculation crosses an interface or persists into a
shared document. The chapter does not insist on universal
discipline; section 1.6 is most strict at interfaces, in handoffs
between systems, and in any quantity that will be read later by
someone who has not seen the calculation. Internal scratch work in
a single notebook session, with the unit system established at the
top of the page, can compress.
The strong version of the position is that the engineer who has
internalised the discipline pays no overhead at all: the unit chain
appears in the engineer's mental write-out automatically. The
overhead is only in the explicit-document version, and the
explicit-document version is cheap insurance.
\end{exercise}

\begin{exercise}
A senior engineer says: ``I do not need a proof; my intuition has
been right for thirty years.'' State the conditions under which
this position is defensible and the conditions under which it is
not. Identify a specific class of system in which intuition built
on past experience reliably fails, even for an experienced engineer.

\paragraph{Discussion.}
Defensible when: the senior engineer is working inside the operating
envelope where the thirty years of experience were accumulated; the
system's behaviour at the edges of that envelope was observed and
remembered; the failure modes the engineer has internalised are the
failure modes the system actually exhibits. The senior engineer's
intuition is a compressed statistical summary of a long working
career; when the working domain is stable, the summary is accurate
and the proof would only confirm it.

Not defensible when: the system has moved outside the envelope the
intuition was built on. A class of system in which intuition
reliably fails: software systems whose state space is exponential in
the number of components, where the failure mode is an interleaving
the engineer has never observed because it occurs with frequency
$10^{-8}$ per operation and the engineer has watched $10^{6}$
operations. The Ariane 5 case is the canonical aerospace version of
this failure: thirty years of Ariane 4 experience did not contain
the Ariane 5 trajectory point at which the overflow occurred. The
senior engineer's intuition was reliable for Ariane 4. It was
silently unreliable for Ariane 5, and the proof discipline that
would have caught the gap was not applied.
\end{exercise}

\begin{exercise}
Identify one notation habit of your own that this chapter has
caused you to reconsider. State the habit, the rule from section
1.6 that it violates, and a concrete change you will make in the
next document you write.

\paragraph{Discussion.}
The exercise has no single correct answer; the reader's submission
stands on the concreteness of the habit named, the precision of the
rule cited, and the testability of the concrete change. A strong
submission names a specific habit (``I use $h$ for both heat-
transfer coefficient and Planck's constant in the same chapter''),
cites the rule (overloading), and commits to a concrete change
(``I will use $h_{\text{c}}$ for heat-transfer coefficient and
reserve $h$ for Planck's constant when both appear''). A weak
submission names the habit vaguely or commits to a vague change.
The discipline of the exercise is the commitment to a change the
reader can audit themselves on, six months later, in their own
next document.
\end{exercise}
