\chapter{Requirements and specification}
\label{vol11:ch02}

\epigraph{The requirement is the contract the design will be judged against; the specification is the contract the manufacturer will be judged against.}{}

\chapmeta{Half-life: medium-life. Archetypes: balance, ethics. Exercise target: 25. Project track: hybrid.}

\noindent
A design is judged against its requirements. The requirements are the engineering encoding of what the design must do; the specification is the engineering encoding of what the design must be. The two are distinct: requirements describe function and performance; specifications describe form and substance. Both are documents that the design team authors, the design organisation commits to, and the manufacturer or builder implements against. The discipline of requirements engineering is the discipline of producing documents that are clear enough, complete enough, and verifiable enough that the design's eventual success can be assessed against them.

\section{Why requirements precede design}\pathtag{core}

A design that proceeds before requirements are established is a design that defines its own success criteria as it goes. The criteria drift to match the design; the design's eventual evaluation against them is circular. The engineering professional record is full of products that "met all their requirements" because the requirements were back-fitted to what was built. The discipline of requirements engineering insists that the requirements be established first, by parties accountable to the user need, and that the design then be evaluated against them.

The functions requirements perform:

\textbf{Contract with the user.} Requirements articulate what the user will receive. The articulation gives the user the basis to accept or reject the deliverable. Without explicit requirements, the user's reaction to the delivered design is unconstrained by prior commitment; the user may legitimately reject a design that the design team considers complete because the user's expectations were different.

\textbf{Contract within the design team.} Requirements coordinate the work of multiple engineers on the same design. Each engineer designs against the shared requirements; the interfaces between subsystems are defined by the requirements; the integration of the subsystems is the assembly against the requirements. Without shared requirements, the subsystems may not integrate.

\textbf{Basis for verification.} Requirements specify what must be verified about the design. The verification activities (analysis, test, demonstration, inspection) are organised against the requirements; each requirement has a verification method and an acceptance criterion. Without requirements, verification is unstructured.

\textbf{Defence against scope creep.} Requirements define the scope of the design. Mid-project additions are explicit changes to the requirements; the scope expansion is visible and can be priced. Without requirements, additions accumulate informally; the eventual scope exceeds the original plan without explicit decision.

\textbf{Communication with regulators and certifiers.} Regulatory frameworks require that designs meet specified requirements; the certification documents reference the requirements explicitly. Without requirements, the regulatory engagement is undefined.

The cost of skipping requirements work is the cost of the rework when the design proves not to address the actual need. The cost is reliably higher than the cost of doing the requirements work in the first place. The discipline of "shift left" (catch requirements issues early, at requirements stage) is the engineering expression of this fact.

\begin{principle}[Requirements before design]
A design that proceeds before its requirements are explicitly stated cannot be judged against them; the design's success criteria drift to match what was built. The discipline of requirements engineering is the discipline of fixing the criteria before the design begins and committing to evaluate the design against the fixed criteria.
\end{principle}

\section{Eliciting requirements}\pathtag{core}

Requirements come from the user, the client, the operating environment, the regulatory framework, and the design organisation itself. Eliciting them is a distinct activity from authoring them: the elicitation is the discovery of what the requirements should be; the authoring is the documentation of what they are.

The elicitation methods:

\textbf{Stakeholder interviews.} The engineer interviews the parties whose interests the design must serve: the user, the client, the operator, the maintainer, the regulator, the manufacturer. The interview is structured to surface the requirement; the structure is the interviewer's discipline. The output is a set of requirement candidates that the requirements document subsequently refines.

\textbf{Use-case analysis.} The engineer characterises the situations in which the design will be used: who is the user; what is the user trying to accomplish; what is the surrounding environment; what is the sequence of operations; what can go wrong. The analysis produces use cases that requirements are derived from. The use cases also feed user research (Chapter 3) and testing (Chapter 8).

\textbf{Operational scenarios.} The engineer walks through complete operational scenarios from the user's perspective: a day in the life; a complete mission; a full production run; an emergency response. The scenarios reveal requirements that point-use-case analysis would miss.

\textbf{Field observation.} The engineer observes the actual operation of the current solution (or a comparable solution if no current solution exists). The observation reveals requirements that interviews and use cases miss because users do not articulate what they take for granted. The "shadowing" technique in user research and the "Gemba walk" in industrial engineering are forms of field observation.

\textbf{Document review.} The engineer reviews existing documentation: the previous-generation system's requirements; the competitor's published specifications; the standards bodies' applicable standards; the regulatory framework. The review identifies requirements that have already been established for the class of system.

\textbf{Workshops.} The engineer (with a facilitator) convenes a workshop of stakeholders to elicit requirements collaboratively. The collaborative format surfaces requirements through interaction; participants build on each other's contributions. The format requires investment in preparation and facilitation; the output is typically higher quality than equivalent investment in individual interviews.

\textbf{Prototype-based elicitation.} The engineer builds a low-fidelity prototype and presents it to stakeholders; the stakeholders' reactions surface requirements that abstract discussion does not. The "Wizard of Oz" prototype (a human pretends to be the system's automated response) is a common variant.

The elicitation must be balanced against itself: too little produces requirements that miss substantive needs; too much produces requirements documents that are too detailed to act on. The judgment of "enough" depends on the design's stakes, the maturity of the application domain, the experience of the engineering team, and the cost of late-discovered requirements changes.

\section{Functional vs non-functional requirements}\pathtag{core}

Requirements divide along two axes:

\textbf{Functional requirements} describe what the design does. "The system shall measure ambient temperature to $\pm 0.5\,\degC$ accuracy over the range $-20$ to $+50\,\degC$." "The vehicle shall accelerate from 0 to $100\,\text{km/h}$ in $\leq 12\,\text{s}$." "The user shall be able to complete the registration process in $\leq 5$ minutes." Functional requirements specify behaviour; they are typically the easier requirements to elicit, document, and verify.

\textbf{Non-functional requirements} describe properties the design must have. Reliability ("the system shall achieve mean time between failures of $\geq 50{,}000$ operating hours"); maintainability ("the system shall be repairable by a Level-2 technician in $\leq 2$ hours"); usability ("a novice user shall be able to complete the basic operation without training"); security ("the system shall protect user credentials according to NIST SP 800-63"); safety ("the system shall fail to a safe state on detected fault"); cost ("the unit production cost shall be $\leq \$200$"); weight; size; power consumption; environmental conditions of operation; lifecycle considerations (manufacturability, sustainability, accessibility).

Non-functional requirements are typically harder to write than functional ones. "The system shall be reliable" is not a requirement; "the system shall achieve MTBF of $\geq 50{,}000$ hours under operating profile X with 90\% confidence over a 3-year service period" is a requirement. The discipline of writing non-functional requirements is the discipline of finding the measurable quantity that captures the quality the stakeholder is asking for.

Non-functional requirements also interact more strongly. A requirement for low cost may be in tension with a requirement for high reliability; a requirement for high security may be in tension with a requirement for easy access; a requirement for small size may be in tension with a requirement for thermal headroom. The requirements document captures the tensions; the design resolves them by allocating margins (Chapter 5).

Some requirements categories that recur across domains:

\textbf{Performance.} Speed, throughput, response time, accuracy, range, capacity.

\textbf{Reliability.} MTBF, MTTR, availability, fault tolerance, fail-safe behaviour.

\textbf{Safety.} Hazard exposure limits, protection requirements, emergency response.

\textbf{Security.} Authentication, authorisation, confidentiality, integrity, availability, audit.

\textbf{Usability.} Time-on-task, error rate, satisfaction, accessibility.

\textbf{Maintainability.} MTTR, diagnostics, parts availability, service intervals.

\textbf{Manufacturability.} Process compatibility, tolerance, assembly time, test access.

\textbf{Environmental.} Operating temperature, humidity, vibration, shock, EMI, ingress protection.

\textbf{Lifecycle.} Service life, disposal, recyclability, environmental impact.

\textbf{Cost.} Unit cost, lifecycle cost, NRE, total cost of ownership.

\textbf{Schedule.} Delivery date, milestone dates, time-to-market.

\textbf{Regulatory.} Applicable standards, certifications required, jurisdictions covered.

\section{Specifications: how precise is precise enough}\pathtag{core}

The specification translates the requirements into engineering description: dimensions, materials, tolerances, interfaces, test procedures. The specification is what the manufacturer uses to build the design; what the inspector uses to verify the build; what the customer uses to accept delivery. The precision of the specification determines what gets built.

The precision principle: specify only what matters; specify nothing else. Over-specification (specifying details that do not affect the design's function) constrains the manufacturer unnecessarily, raises costs, and reduces design flexibility. Under-specification (omitting details that do affect the function) produces variation in what gets built. The judgment of what to specify is the substantive design judgment captured in the specification document.

The dimensions of specification precision:

\textbf{Dimensions and tolerances.} The geometry is specified to the precision required for function. A clearance fit between two parts requires the dimensions of both parts plus the tolerance on each; the tolerance stack-up (Chapter 5, and Volume II Chapter 14 on tolerance analysis) determines what precision is required. Over-tight tolerances make manufacturing expensive; over-loose tolerances make assembly unreliable.

\textbf{Materials.} The materials are specified to the level required for the function. A structural part may require a specific alloy with specific heat treatment; a non-structural part may accept any material in a broad class. The specification balances the design's requirements against the available materials' cost and processability.

\textbf{Surface finish.} The surface finish (roughness, waviness, lay) is specified where it matters for function (sealing surfaces; sliding surfaces; aesthetic surfaces) and left unspecified where it does not.

\textbf{Functional behaviour.} The behaviour the design must exhibit is specified to the level the verification will measure. A control system's response time may be specified as a step-response settling time; a sensor's accuracy as a maximum error over a specified input range; a communication interface as a protocol specification.

\textbf{Interfaces.} The interfaces to other systems are specified: mechanical (mating geometry, alignment, retention); electrical (voltage, current, protocol, connector); thermal (heat transfer; envelope); informational (data formats, protocols, latency). Interface specifications are the points at which the design's success depends on agreement with the other side of the interface.

\textbf{Test procedures.} The procedures used to verify the design's behaviour are specified: test setup, test inputs, expected outputs, acceptance criteria, test conditions. Test procedure specification is what makes the verification reproducible and what makes the design's acceptance decidable.

The specification document is the contractual basis for what gets built. The discipline of authoring is the discipline of being precise where precision matters and silent where it does not; the discipline of using is the discipline of building exactly what is specified and asking for clarification when the specification is ambiguous.

\begin{archetype}[balance]
Requirements and specifications are the balance archetype's expression in the discipline of design. The engineer balances the user's want against the constraints; the design's flexibility against the user's certainty; the specification's precision against the manufacturing cost; the document's completeness against the document's usability. The balance is not in any specific requirement; it is in the document as a whole.
\end{archetype}

\section{Requirements traceability}\pathtag{standard}

Traceability is the discipline by which every requirement is linked to its source (the user need or stakeholder input that gave rise to it) and to its consequence (the design element or test that addresses it). The traceability is documented in a traceability matrix or in a database; the discipline is maintained throughout the project lifecycle.

The traceability links serve several functions:

\textbf{Change impact analysis.} When a requirement changes, the traceability identifies the design elements that are affected. The change-control process uses the traceability to scope the changes; without traceability, the scope of change is opaque and changes propagate as surprises.

\textbf{Verification coverage.} The traceability identifies, for each requirement, the verification method(s) that address it. The coverage analysis identifies requirements that lack verification (a gap that the test plan must address) and verifications that do not address a requirement (a test that may be unnecessary).

\textbf{Stakeholder satisfaction.} The traceability allows stakeholders to verify that their inputs are reflected in the design. A stakeholder who asked for X can follow the traceability from "X" to the requirement, to the design element, to the verification that confirms X is met.

\textbf{Regulatory compliance.} Regulators require, in many industries, that the design's compliance with regulations be traceable from the regulation to the design to the test evidence. The traceability is part of the certification submission.

\textbf{Forensic analysis.} If the design fails in service, the traceability is part of the forensic analysis (Chapter 12 of Volume X): which requirement was inadequate; which design element failed against the requirement; which verification missed the failure mode.

The traceability is maintained in tools (Doors, Polarion, Jama, the lower-end JIRA-based workflows). The tool's discipline is what determines whether the traceability is current; tools that require manual maintenance of links typically degrade over time as the project changes; tools that auto-generate links from project artefacts (model-based systems engineering tools) are more reliable but require investment in the underlying models.

\section{Verification and validation distinguished}\pathtag{core}

Verification and validation are two distinct concepts often conflated. Verification asks: "Did we build the design right?" Validation asks: "Did we build the right design?"

\textbf{Verification} is the activity of confirming that the design meets its requirements. The requirements specify the design's expected behaviour; the verification activities (analysis, test, demonstration, inspection) confirm that the actual design exhibits the expected behaviour. Verification is engineering-internal: it asks whether the design meets its own specification.

\textbf{Validation} is the activity of confirming that the design serves the user's need. The user need was the starting point of the design process (Chapter 1); the validation activities (user testing, field trials, market response) confirm that the actual design serves the actual need. Validation is engineering-external: it asks whether the design meets the user's purpose.

The distinction matters because designs can verify successfully and validate unsuccessfully. The design met every requirement; the user does not want the result. The failure is at the requirements stage (the requirements did not capture the user need adequately), not at the design or build stage. The remediation is re-specification, not re-design.

The converse can also occur: designs can validate well even when verification falls short on some specific requirements. The user is happy; the design did not meet every requirement. The failure (such as it is) is at the requirements stage in a different sense: the requirement may have been a misjudgement about what the user needed. The remediation is to re-think the requirement.

The two activities are performed at different stages and by different methods:

\textbf{Verification methods.} Analysis (compute the design's expected behaviour from the design specification); test (operate the design under controlled conditions and measure the behaviour); demonstration (operate the design under intended-use conditions and observe the behaviour); inspection (examine the design for compliance with specification, often visually or with measurement). Each method has appropriate domains; the verification plan selects methods per requirement.

\textbf{Validation methods.} User testing (real users operate the design under realistic conditions; the user's actual behaviour and feedback are observed); field trials (the design is deployed in operational settings and operational data are collected); pilot deployment (the design is deployed at limited scale before broad rollout); market evaluation (the design is launched and adoption, satisfaction, and continued use are measured). Validation is typically more expensive than verification and more difficult to plan; the discipline is to identify the validation evidence the design needs before release and to plan its collection.

The verification-validation distinction is foundational to systems engineering (the V-model is named for the verification-validation V it represents); it is taught explicitly in the major standards (ISO/IEC/IEEE 15288 for systems and software engineering; INCOSE handbook).

\section{Requirements as a conflict-management tool}\pathtag{standard}

Multiple stakeholders typically have incompatible requirements. The user wants high performance and low cost; the manufacturer wants high tolerance to manufacturing variation and the user wants tight performance; the regulator wants exhaustive testing and the schedule wants rapid delivery. The requirements document is the formal articulation of these conflicts and the basis for their resolution.

The conflict-management mechanisms:

\textbf{Priority assignment.} Each requirement is assigned a priority (must, should, could, will-not, in the MoSCoW framework; alternatives include simple high-medium-low, or numeric scoring). Priorities determine which requirements are first to be cut when constraints force tradeoffs.

\textbf{Acceptance ranges.} Requirements are specified with target values and acceptance ranges rather than single points. The acceptance range provides design flexibility; the target value indicates the preferred value within the range.

\textbf{Conditional requirements.} Requirements are made conditional on operating modes or contexts. "Under normal operation, the response time shall be $\leq 100\,\text{ms}$; under degraded mode, the response time shall be $\leq 1\,\text{s}$." Conditional requirements allow the design to address different conditions appropriately.

\textbf{Negotiated requirements.} The requirements are explicitly negotiated among stakeholders; the negotiation produces compromises that all parties commit to. The negotiation is the engineering equivalent of contract negotiation; the result is the contract.

\textbf{Phased requirements.} Some requirements are deferred to later releases or phases; the initial design addresses a subset, with the remaining requirements addressed in subsequent iterations. The phasing requires explicit articulation of what is in the initial scope and what is deferred.

\textbf{Escalation paths.} Disputes about requirements that cannot be resolved at the project level escalate to designated higher levels (program manager; engineering director; product committee). The escalation paths are documented; the levels at which different classes of dispute are resolved are explicit.

The conflict-management value of requirements documents is among their highest practical functions. A project without explicit requirements has conflicts that are equally real but that surface as opaque political fights; a project with explicit requirements channels the conflicts through documented mechanisms with explicit decisions and accountability.

\section{Failure: missed, ambiguous, or self-contradictory requirements}\pathtag{core}

The failures of requirements engineering produce predictable design outcomes:

\textbf{Missed requirements.} A requirement that should have been captured was not. The design that proceeds without it may meet all the captured requirements while failing in service because the un-captured requirement is violated. The Mars Climate Orbiter failure (Volume X Chapter 5) is partly a missed-requirement story: the requirement that all interface units be metric was not explicitly captured for the navigation software interface; the contractor delivered the software in pound-seconds; the spacecraft entered Mars atmosphere too low and burned up.

\textbf{Ambiguous requirements.} A requirement is stated in language that admits multiple interpretations. The designer interprets it one way; the user expects another; the delivered design satisfies the designer's interpretation but not the user's. Ambiguity is reduced by using measurable, testable language with explicit units, ranges, and conditions; ambiguity is preserved by using natural-language formulations like "the system shall be user-friendly" or "the response shall be timely."

\textbf{Self-contradictory requirements.} Two requirements demand outcomes that cannot both be achieved. The design satisfies one and violates the other; the choice of which to satisfy is implicit (the designer chooses) rather than explicit (the requirements document acknowledged the conflict and resolved it). The discipline of requirements review includes the discipline of checking for self-contradiction; the check is non-trivial in documents with hundreds or thousands of requirements.

\textbf{Untestable requirements.} A requirement is stated in language that no verification activity can decisively confirm or refute. "The system shall be intuitive to use" admits no test; "a novice user shall complete the registration task in $\leq 5$ minutes with no assistance, with success rate $\geq 90\%$" admits a test. Untestable requirements are not requirements; they are aspirations that the requirements document failed to translate into engineering input.

\textbf{Over-constrained requirements.} The requirements specify a particular design rather than the function the design must serve. "The system shall use a Type-X widget" is over-constrained if any widget that performs the function would do; the over-constraint limits the design space without engineering justification. The discipline of requirements writing is to specify what the design must do, not how it must do it.

\textbf{Stale requirements.} The requirements were captured at the start of the project; the world changed during the project; the requirements no longer reflect the current need. The design satisfies the original requirements and fails the current need. The discipline of requirements maintenance is the discipline of re-engaging with stakeholders periodically and updating the requirements as the need evolves.

The Therac-25 incidents (Volume X Chapter 5) are partly a requirements failure: the software-only interlock requirement that replaced the hardware interlock of the Therac-20 was specified without adequate attention to the software's fault modes; the requirement implicitly assumed software correctness that the verification did not establish.

The 737 MAX MCAS failure (Volume X Chapter 8) is partly a requirements failure: the requirement for MCAS authority was expanded during development without the safety requirements being rerun against the expanded authority; the single-AoA input requirement was inconsistent with the function's effective authority but the requirements review did not catch the inconsistency.

The Bhopal disaster (Volume X Chapter 10) is partly a requirements failure: the safety requirements for the MIC storage and processing systems were either inadequate at design or were eroded in operation; the requirements that should have been binding (refrigeration; flare; scrubber; inventory limits) were not enforced.

The recurring lesson is that requirements failures are foundational: they cannot be remediated by execution discipline at later stages. The discipline of requirements engineering is the discipline of preventing the foundational failure.

\begin{principle}[Requirements are the foundation]
Most consequential design failures have a requirements failure in their causal chain: a need that was not captured; a requirement that was ambiguous; a contradiction that was not resolved; a verification that could not actually verify. The discipline of requirements engineering is the discipline of producing the documents that the rest of the design process depends on. The cost of inadequate requirements is the cost of all the downstream rework and the residual risk of in-service failure.
\end{principle}

\begin{project}[Requirements document with rubric review]
Select one project from the reader's prior work (a design from the projects of earlier chapters in this book; a personal project; a professional project the reader can use). Write a requirements document for the project following the discipline of this chapter. The document must include: stakeholders (who has interests in the design); functional requirements (what the design does); non-functional requirements (what properties the design has); priorities (MoSCoW or equivalent); acceptance criteria (how each requirement will be verified); traceability links (each requirement linked to its source). Then evaluate the document against a rubric drawn from this chapter: are requirements measurable? unambiguous? consistent? complete? traceable? testable? Deliverable: the requirements document plus a one-page self-review against the rubric.

\textbf{Hazard.} The exercise often reveals that the project's actual requirements were never explicit; the discipline of writing them after the fact is a useful diagnostic for what was implicitly assumed. The point is the discipline of explicit articulation, not the production of a beautiful document.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A requirements document has 50 functional requirements and 30 non-functional requirements. Each requirement requires (on average) 1 hour to elicit, 2 hours to write, 1 hour to review, and 3 hours of verification effort. Compute the total requirements-engineering effort in person-hours. Compare to the typical effort allocated in a project with this requirement count.
\end{exercise}

\begin{exercise}
For a project with $n$ requirements, the verification effort is proportional to $n$ and the design effort to verify against the requirements is proportional to $n^{1.5}$ (because of inter-requirement interactions). Compute the ratio of verification to design effort for $n = 10$, $n = 100$, and $n = 1000$. Identify the requirement-count regime in which verification dominates and the regime in which design dominates.
\end{exercise}

\begin{exercise}
A requirements traceability matrix links $R$ requirements to $D$ design elements to $V$ verification activities. Compute the number of cells in a 3-dimensional matrix and the storage required for a sparse representation if average requirement links to 2 design elements and 1 verification activity. Compare for $R, D, V = 100, 1000, 10000$.
\end{exercise}

\begin{exercise}
The acceptance criterion for a requirement is "$\geq 90\%$ success rate over a sample of $n$ trials." Compute the minimum $n$ such that, if the true success rate is 95\%, the observed sample produces $\geq 90\%$ success with 95\% confidence. Discuss the design implications.
\end{exercise}

\begin{exercise}
A design has 200 requirements with priorities distributed as 40\% must, 30\% should, 20\% could, 10\% will-not. Compute the count in each priority. Discuss the implications for scope management when the project's effort budget covers only 60\% of the original scope.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the conditions under which a requirements set is "complete" in the formal sense: every operationally significant behaviour of the system is constrained by at least one requirement. Identify why completeness in this sense is unachievable in practice and what the working approximation is.
\end{exercise}

\begin{exercise}
For a multi-stakeholder requirements negotiation with $K$ stakeholders each having $N$ ranked requirements, derive an aggregation rule that produces a single ranked requirements list. Identify the conditions under which the aggregation is fair and the conditions under which it is contested.
\end{exercise}

\begin{exercise}
Derive the relationship between requirement ambiguity (variance in interpretation by different readers) and design rework effort. Identify the conditions under which it is more cost-effective to reduce ambiguity at requirements stage versus accept some ambiguity and pay the rework.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the fraction of project effort in a typical engineering project devoted to requirements work versus design, prototyping, testing, and integration. Compare across project types (research; product development; safety-critical; software).
\end{exercise}

\begin{exercise}
Estimate the cost of a requirements change discovered at: requirements stage; concept stage; detailed design; prototype; production; field deployment. Identify the cost-ratio range across stages.
\end{exercise}

\begin{exercise}
Estimate the fraction of accident-investigation reports (NTSB, AAIB, BEA, CSB) that identify "requirements failure" (missed, ambiguous, contradictory, untestable) as a contributing cause. Compare to the fraction identifying "execution failure" (the design met requirements but the implementation was flawed).
\end{exercise}

\begin{exercise}
Estimate the time required to write a "good" requirement (specific, measurable, testable, unambiguous) versus a "weak" requirement (vague, untestable, ambiguous). Identify the project-effort implication of insisting on good requirements.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Implement a simulation of requirements elicitation in which each elicitation method (interview, use case, workshop, observation, document review) has a per-requirement discovery probability and a per-interview cost. Sweep the elicitation budget; identify the optimal allocation of effort across methods.
\end{exercise}

\begin{exercise}
Simulate a requirements traceability matrix's degradation over time under a model of project changes (10\% of requirements change per quarter; 50\% of changes are reflected in the traceability; 50\% are not). Compute the matrix's accuracy after 1, 2, and 3 years.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a requirements-document template for a small engineering project (e.g., a household device). Include sections for stakeholders, functional requirements, non-functional requirements, constraints, assumptions, verification approach, and traceability. Justify each section's necessity for projects at this scale.
\end{exercise}

\begin{exercise}
Design a requirements-review protocol for an engineering team. Specify the participants, the artefacts reviewed, the questions asked at each pass, and the criteria for the review to pass.
\end{exercise}

\begin{exercise}
Design a change-control process for requirements changes during design. Specify the triggers, the assessment, the decision authority, and the documentation. Identify how the process scales from minor wording changes to substantive requirement additions.
\end{exercise}

\begin{exercise}
Design a verification plan for one specific non-functional requirement (e.g., "the system shall achieve MTBF of $\geq 10{,}000$ hours under operating profile X"). Specify the verification method, the test setup, the test conditions, the acceptance criteria, and the resources required.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: the Mars Climate Orbiter failure (Volume X Chapter 5) is principally a requirements-engineering failure rather than a software-engineering failure or an interface-discipline failure. Identify the requirements-document moves that would have prevented the failure and the costs of those moves.
\end{exercise}

\begin{exercise}
Discuss the verification-validation distinction. Argue: an engineering organisation that conflates the two will systematically deliver designs that verify well and validate poorly. Identify the structural conditions that produce the conflation and the institutional changes that would address them.
\end{exercise}

\begin{exercise}
Argue: requirements documents that are too detailed are as harmful as requirements documents that are too sparse. Identify the failure modes of each extreme and discuss what determines the appropriate level of detail for a given project.
\end{exercise}

\begin{exercise}
Discuss the role of regulatory requirements in design. Argue: regulatory requirements are constraints on design rather than inputs to design. Argue the opposite. Identify the design-process implications of each position.
\end{exercise}

\begin{exercise}
A safety-critical design has a requirement that conflicts with a user-experience requirement. The safety requirement comes from the regulator; the user-experience requirement comes from the marketing department. Discuss the engineer's options for resolving the conflict and the professional-ethics framework that bears on the resolution.
\end{exercise}

\begin{exercise}
Consider the requirement "the system shall be user-friendly." Critique the requirement against the discipline of this chapter. Rewrite it as one or more measurable, testable requirements. Discuss what user-research evidence would be needed to set the acceptance thresholds.
\end{exercise}

\begin{exercise}
Take one safety-critical artefact the reader has interacted with (an aircraft, an elevator, a medical device, an infrastructure component). Identify three requirements (functional or non-functional) that the artefact's design must have addressed. For each, identify the verification method that would establish compliance and the consequence of failure. Discuss whether the verification methods you identified are likely the ones actually used in the artefact's certification.
\end{exercise}
