\chapter{Instabilities, runaway, buckling, resonance}
\label{vol10:ch04}

\epigraph{Stability is the property; instability is the engineer's nightmare.}{}

\chapmeta{Half-life: durable. Archetypes: failure, coupling. Exercise target: 25. Project track: simulation.}

\noindent
Instability is the failure mode in which a system's equilibrium gives way to runaway or oscillation. The mechanisms span engineering: a strut buckles under axial load; a chemical reaction runs away to explosion; a bridge resonates with wind into destruction; a control loop self-oscillates into damage. Each instability has its own physics; they share the structural feature that small perturbations grow rather than decay. This chapter develops buckling at engineering depth, aeroelastic instability, thermal and chemical runaway, mechanical resonance failures, numerical instability as engineering failure, and the major case studies.

\section{Buckling at depth}\pathtag{standard}

Buckling is the loss of geometric stability under compressive load: a slender column, a thin plate, a thin shell, or a frame can fail at loads well below the material's compressive strength when the deformed configuration is preferred to the undeformed.

Euler buckling of a column: the critical load is $P_{cr} = \pi^2 E I / L_e^2$, where $E$ is Young's modulus, $I$ is the cross-section's moment of inertia, $L_e$ is the effective length (function of end conditions: pinned-pinned $L_e = L$, fixed-free $L_e = 2L$, fixed-fixed $L_e = 0.5L$, fixed-pinned $L_e \approx 0.7L$). The Euler formula applies for slender columns where the buckling occurs before material yielding; the slenderness ratio $L_e/r$ (with $r$ the radius of gyration) separates slender from intermediate behaviour.

Intermediate columns fail by inelastic buckling: the column yields locally and then buckles in the yielded configuration. The Johnson formula or the AISC column equations cover this regime. Short columns fail by crushing rather than buckling.

Plate buckling: thin plates in compression buckle at $\sigma_{cr} = k \pi^2 E / 12(1-\nu^2)(b/t)^2$, with $b/t$ the width-to-thickness ratio, $\nu$ Poisson's ratio, and $k$ a coefficient depending on boundary conditions. Post-buckling, plates can carry additional load (the effective-width concept).

Shell buckling: thin cylindrical shells under axial compression have the theoretical critical stress $\sigma_{cr} \approx 0.6 E t / R$, but real shells fail at $25\%$ to $50\%$ of this due to imperfection sensitivity. Spherical shells under external pressure similarly fail well below theoretical. The shell-buckling literature emphasises imperfection sensitivity; design codes (ASME, Eurocode) include large knockdown factors.

Frame buckling: the full frame's stability depends on the interaction between members; the analysis combines individual member buckling with the system's lateral stiffness. Sway frames (no lateral restraint) are particularly vulnerable.

Lateral-torsional buckling: a beam in bending can buckle laterally if not braced; the deep, narrow beam under load is the canonical case. Bracing reduces the effective length for lateral-torsional behaviour.

\begin{archetype}[failure: as instability past critical load]
Buckling and other instabilities share the structural feature that the system has a stable configuration up to a critical load, and an unstable (or alternative-stable) configuration beyond. The engineering challenge is identifying the critical load, designing with margin, recognising the imperfection sensitivity that reduces real-world capacity, and detecting the precursors before catastrophic failure. The discipline generalises beyond mechanical buckling to thermal runaway, chemical runaway, control-system instability, and other instability modes.
\end{archetype}

\section{Aeroelastic instability}\pathtag{mastery}

Aeroelasticity is the interaction of aerodynamic forces with structural deformation. The structure deforms under aerodynamic load; the deformation changes the aerodynamic load; if the feedback is destabilising, the system enters limit-cycle oscillation or unbounded growth.

Standard aeroelastic phenomena:

\textbf{Divergence:} the static deformation under aerodynamic load increases the aerodynamic load. Above the divergence speed, the structure deforms without bound. Wing torsional divergence is the classical example: increasing angle of attack increases lift, which increases torque, which increases angle of attack. The divergence speed is calculable from the wing's torsional stiffness and aerodynamic characteristics.

\textbf{Flutter:} dynamic instability in which two or more modes couple via aerodynamic forces to produce sustained oscillation. The classical bending-torsion flutter of an aircraft wing: above the flutter speed, the energy input from the airflow exceeds the energy dissipated by the structure; the oscillation grows. Flutter is the major aerospace structural concern; flight tests verify the flutter margin before the aircraft is cleared to fly the operational envelope.

\textbf{Stall flutter:} oscillation in the stalled flow regime; rotor blades and propellers can experience stall flutter at certain operating conditions.

\textbf{Galloping:} bluff cross-sections (square, rectangular, ice-coated wires) can experience cross-flow instability in steady wind; the negative aerodynamic damping produces large-amplitude oscillation.

\textbf{Vortex-induced vibration (VIV):} alternating vortex shedding from a cylinder in cross-flow produces oscillatory force; if the shedding frequency matches a structural natural frequency, lock-in occurs and the response is amplified. Tall stacks, chimneys, and offshore risers experience VIV; mitigation includes vortex spoilers (helical strakes), increased damping, or stiffness changes.

\textbf{Buffeting:} response to turbulent or impulsive aerodynamic excitation; not strictly an instability but can produce large stress amplitudes.

The Tacoma Narrows Bridge collapse (\cite{acc:tacoma-narrows-1941}, 1940) was an aeroelastic failure: the bridge deck's torsional mode was excited by the wind, producing self-excited oscillation that grew in amplitude until collapse. The mechanism (sometimes simplified as "resonance with vortex shedding") is actually negative aerodynamic damping in the torsional mode at the operating wind speed. The accident catalysed the discipline of aeroelastic analysis for civil structures.

\section{Thermal runaway}\pathtag{standard}

Thermal runaway is the instability in which a temperature increase causes increased heat generation, which produces further temperature increase, until material limits are exceeded and catastrophic damage occurs.

Classical examples:

\textbf{Semiconductor thermal runaway:} a transistor's reverse-saturation current doubles every $10\,^\circ\text{C}$. In a circuit with positive thermal-electrical feedback (e.g., bipolar transistor with poor thermal stability), increased temperature causes increased current, which produces more heat. The transistor can destroy itself. Mitigation: thermal-stabilising circuits (emitter degeneration), thermal coupling between paired devices, heatsink design.

\textbf{Lithium-ion battery thermal runaway:} elevated temperature accelerates the cell's exothermic reactions; the heat generation accelerates; the cell vents or burns; adjacent cells in the pack are heated and may follow. The 2013 Boeing 787 battery incidents (\cite{acc:ntsb-aar14-01-787-battery}) involved thermal runaway. Mitigation: cell-level temperature monitoring, thermal isolation between cells, conservative charging, fire-resistant pack designs.

\textbf{Nuclear thermal runaway:} the reactor's power varies with temperature through the temperature coefficient of reactivity. A positive coefficient (rare in modern designs) produces a thermal-runaway hazard. The 1986 Chernobyl accident (\cite{acc:iaea-insag7-1992}) involved a positive void coefficient at low power; reactivity increased as cooling water voided; the reactor power increased; the steam-explosion damage destroyed the reactor.

\textbf{Chemical thermal runaway:} exothermic reactions produce heat; the reaction rate accelerates with temperature; the heat-removal capacity may be exceeded. The 1984 Bhopal disaster (\cite{acc:icmr-bhopal-2004}) involved water entering a methyl-isocyanate storage tank; the resulting exothermic reaction generated heat and gas faster than the cooling system could remove it; the relief system was inadequate; the tank vented toxic gas to the surroundings.

\textbf{Internal combustion engine thermal runaway:} pre-ignition in a spark-ignition engine produces excess heat; if continued, the engine can be destroyed. Detonation (knock) is similar but smaller-scale. Mitigation: fuel-octane control, ignition-timing control, cooling-system design.

Thermal runaway requires positive feedback (temperature increase produces heat-generation increase) and inadequate heat-removal capacity. The engineering response: identify the runaway potential during design; provide heat removal sized for the worst-case generation; provide detection and shutdown (overtemperature trips, relief valves, emergency cooling); provide post-event protection (containment, blast-resistant design).

\section{Chemical runaway}\pathtag{standard}

Chemical runaway extends thermal runaway to the broader class of self-accelerating chemical reactions. The mechanisms:

\textbf{Exothermic chemistry:} the reaction releases heat; the heat accelerates the reaction (Arrhenius); positive feedback. Most thermal-runaway examples are also chemical-runaway examples.

\textbf{Autocatalysis:} a reaction product catalyses the reaction; the rate accelerates as products accumulate. Some polymerisation reactions and certain decomposition reactions.

\textbf{Free-radical chain reactions:} initiation produces radicals; propagation reactions multiply the radicals; the rate accelerates explosively. Combustion, polymerisation, atmospheric chemistry.

\textbf{Pressurised systems:} reactions in closed vessels produce pressure rise; the pressure increases the reaction rate (through partial-pressure effects or vessel temperature); positive feedback.

The Bhopal disaster combined exothermic chemistry with inadequate cooling, suppressed alarms, deferred maintenance, and inadequate emergency response. The reaction itself was well-understood; the system's lack of margin (cooling, instrumentation, relief, containment) was the failure mode.

Process-safety engineering addresses chemical runaway systematically:

\textbf{Reaction-hazard characterisation:} the reaction's heat-of-reaction, rate constants, decomposition products, and pressure generation are measured (DSC, ARC, vent-sizing tests). The data informs vessel design.

\textbf{Vent and relief design:} the worst-case scenario's pressure-rise rate is computed; the relief device is sized to handle the resulting flow without exceeding the vessel's pressure limit. The discipline is the Design Institute for Emergency Relief Systems (DIERS, AIChE) methodology.

\textbf{Layer-of-protection analysis (LOPA):} the protective layers (process design, alarms, safety-instrumented systems, relief, containment) are documented and their performance verified. The methodology informs safety-integrity-level (SIL) requirements.

\textbf{Inherently safer design:} the process is designed to reduce the worst-case consequence: reduce inventory, lower temperature/pressure, use less hazardous chemistry. The principle (Kletz 1978) is the most reliable mitigation: if the hazardous condition cannot occur, no protection is needed.

\section{Mechanical resonance failures}\pathtag{standard}

Resonance failure occurs when an excitation frequency matches a structural natural frequency; the response amplifies and stresses exceed limits.

Standard cases:

\textbf{Forced resonance:} an external excitation drives the structure at its natural frequency. Wind on bridges (Tacoma Narrows), foot traffic on pedestrian bridges (Millennium Footbridge 2000, \cite{acc:dallard-millennium-2001}), reciprocating machinery on supporting structures, rotating-equipment imbalance on bearings.

\textbf{Self-excited oscillation:} the structure's motion produces the excitation (the wind-flow case is aeroelastic; mechanical analogues include some lathe-chatter cases, brake squeal, rotor whirl).

\textbf{Parametric resonance:} time-varying stiffness or geometry produces resonance at frequencies related to the variation. Ship parametric rolling, gear-tooth vibration, certain shaft-vibration modes.

\textbf{Sub-synchronous resonance:} in electrical-mechanical systems (turbine-generator on the grid), the mechanical natural frequency can couple with electrical resonances at lower-than-mains frequencies; the resulting oscillation has destroyed turbine shafts.

The engineering response:

\textbf{Frequency separation:} design the structural natural frequencies away from the excitation frequencies. The standard rule is a factor of $2$ separation; tighter margins demand careful analysis.

\textbf{Damping:} increase the structural damping to reduce the resonance amplification. Tuned mass dampers (e.g., Taipei 101's $660$-ton tuned damper) are the engineering response when frequency separation is not feasible.

\textbf{Isolation:} mechanically isolate the structure from the excitation. Vibration mounts for machinery, base isolation for buildings in earthquake zones.

\textbf{Active control:} sensors and actuators feedback to suppress resonance. Active vibration control is mature for high-performance applications (precision machines, aerospace structures).

\textbf{Avoidance:} restrict operation to conditions that do not excite the resonance. Aircraft flutter margins, rotor critical-speed avoidance.

\section{Numerical instability as engineering failure}\pathtag{mastery}

Numerical instability is the loss of accuracy or stability in a numerical algorithm. The engineering consequence: the simulation, control loop, or analysis produces wrong answers that may be undetected.

Standard mechanisms:

\textbf{Loss of significant digits:} subtracting two nearly-equal floating-point numbers loses precision; the resulting small difference is dominated by round-off error. Accumulated loss across many operations can produce wildly wrong answers.

\textbf{Stiff differential equations:} explicit time-stepping requires step size much smaller than the fastest dynamic; the step-size requirement makes the simulation impractically slow. Implicit methods address this but may introduce numerical damping.

\textbf{Ill-conditioned matrices:} small input changes produce large output changes; the matrix's condition number quantifies the amplification. Inversion of ill-conditioned matrices produces unreliable answers; SVD or pseudoinverse may be needed.

\textbf{Algorithm instability:} the algorithm amplifies round-off rather than averaging it out. Gram-Schmidt orthogonalisation (vs modified Gram-Schmidt) is a classical case.

\textbf{Control-loop limit cycles:} the discrete-time control loop's quantisation produces sustained oscillation at low amplitudes; benign in some applications, problematic in precision.

The Patriot missile clock-drift incident (\cite{acc:gao-patriot-1992}, 1991) was a numerical-precision failure: the system's time was stored as a multiple of $0.1$ seconds; $0.1$ has no exact binary representation; over $100$ hours of operation, the accumulated error was $0.34$ seconds; the missile-tracking radar's range gate was offset; the system missed a Scud missile; $28$ soldiers died. The fix was a known patch in the field that had not been installed. The mechanism is a textbook example of accumulated floating-point error producing operational failure.

Mitigation: use stable algorithms (forward error analysis), use appropriate data types (double precision rather than single where precision matters), monitor numerical health (condition numbers, residuals), validate against analytical or experimental results.

\section{Case: Tacoma Narrows, Texas City refinery 2005, Therac-25 race condition, Lyon-Saint-Exup\'ery station roof}\pathtag{core}

Four instability case studies:

\textbf{Tacoma Narrows Bridge (1940):} as introduced in the aeroelastic section. The torsional flutter is the canonical aeroelastic-instability case. The engineering response shaped wind engineering for civil structures.

\textbf{Texas City refinery explosion (2005):} the BP Texas City refinery raffinate distillation tower exploded during a startup, killing $15$ and injuring over $170$ (\cite{acc:csb-bp-texas-city-2007}). The mechanism involved overfilling of the distillation tower with hot liquid that flashed into vapour when the relief valve opened; the vapour cloud ignited; the explosion damaged surrounding facilities. The accident combined chemical-process hazards (highly flammable hydrocarbons), procedural failures (deferred safety upgrades, inadequate operator training, distracted management), and design failures (relief venting to atmosphere rather than to a flare, inadequate instrumentation). The investigation produced extensive recommendations on process safety, organisational culture, and regulatory enforcement.

\textbf{Therac-25 race condition (1985-1987):} the radiation-therapy machine (\cite{paper:leveson-turner1993}) had a software race condition: the operator's input sequence could leave the machine in an inconsistent state (electron-beam current selected, X-ray-mode shielding not in place); the patient received massive radiation overdose. Six known overdoses, three fatal. The mechanism is software instability: the system's state machine had paths that could be traversed in an order the designers had not anticipated; the hardware interlocks that had prevented the failure in earlier machine generations had been removed in the cost-reduced Therac-25. The case is the canonical reference for software safety in cyber-physical systems.

\textbf{Lyon-Saint-Exup\'ery TGV station roof (2004):} a section of the architecturally distinctive Calatrava-designed roof at the Lyon TGV station collapsed during construction or under load; the analysis identified that the slender steel ribs were vulnerable to buckling under load combinations not adequately addressed in the original design. The case (and similar Calatrava-design incidents) illustrate that aesthetic-led design must still respect buckling and stability margins; the slender forms are also imperfection-sensitive.

The common pattern: the system's instability was foreseeable in principle; the specific combination of conditions (cold weather, operator sequence, design margin, load combination) was at or beyond the qualified envelope; the consequences were severe. The engineering discipline is recognising the instability potential during design, providing margin against the recognised modes, and monitoring or restricting operation to stay within the qualified envelope.

\begin{principle}[The instability-foresight principle]
Instabilities are foreseeable. The mechanism (buckling, aeroelasticity, thermal runaway, chemical runaway, mechanical resonance, numerical instability) is identifiable during design. The engineering discipline is to identify the candidate instabilities, characterise the critical conditions, design with margin against the critical conditions, monitor or restrict operation to stay within the qualified envelope, and provide secondary protection (relief, isolation, automatic shutdown) for cases where the primary protection fails. The instability that is encountered without prior identification is the engineering failure; the instability that is identified, characterised, and managed is the engineering work.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Simulate two instabilities; verify against published accident data]
\textbf{Objective.} Simulate two of the instability mechanisms covered in this chapter (buckling, aeroelasticity, thermal runaway, chemical runaway, resonance, numerical instability); verify your model against published accident data.

\textbf{Method.} Choose two mechanisms. For each: identify a published accident; obtain the relevant parameters (geometry, material, environment, operating conditions); build a simulation appropriate to the mechanism (FEM for buckling, time-domain simulation for runaway, modal analysis for resonance); compare the simulation's prediction to the accident's reported behaviour. Identify what the simulation captures, what it misses, and the engineering lessons.

\textbf{Deliverables.} The two mechanisms with chosen accident cases; the simulation code or model description; the parameters with sources; the simulation results compared to accident data; a written analysis of the model's strengths, gaps, and lessons.

\textbf{Hazard.} Simulation only; the engineering value is in understanding the instability physics.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A pinned-pinned steel column has $L = 3\,\text{m}$, circular cross-section diameter $50\,\text{mm}$. Compute the Euler critical load. State whether the column will yield or buckle first ($\sigma_y = 250\,\text{MPa}$).
\end{exercise}

\begin{exercise}
A thin-walled circular cylinder has $R = 1\,\text{m}$, $t = 5\,\text{mm}$. Compute the theoretical and practical (with knockdown factor $0.3$) axial buckling stresses.
\end{exercise}

\begin{exercise}
A circuit board has natural frequency $80\,\text{Hz}$. The vibration environment has predominant content at $50$ and $100\,\text{Hz}$. State whether resonance is a concern and the engineering action.
\end{exercise}

\begin{exercise}
A reaction has heat-of-reaction $\Delta H = -100\,\text{kJ/mol}$ and rate constant doubling every $10\,^\circ\text{C}$. The reactor cooling can remove $50\,\text{kW}$ at design temperature; the heat-generation at design temperature is $20\,\text{kW}$. Compute the maximum temperature at which thermal balance is maintained.
\end{exercise}

\begin{exercise}
A wind-loaded tall stack has Strouhal-number-derived shedding frequency at the design wind speed of $1.5\,\text{Hz}$. The structural natural frequency is $1.6\,\text{Hz}$. State the engineering concern.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the Euler buckling formula from the equilibrium equation of the deformed column.
\end{exercise}

\begin{exercise}
Derive the criterion for thermal runaway in a CSTR with first-order exothermic reaction and constant cooling.
\end{exercise}

\begin{exercise}
Derive the lock-in condition for vortex-induced vibration of a cylinder in cross-flow.
\end{exercise}

\begin{exercise}
Derive the flutter critical speed for a two-DOF bending-torsion wing model.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the effective length factor for a column with one end fixed in a stiff foundation and the other end pinned to a rigid floor.
\end{exercise}

\begin{exercise}
Estimate the thermal runaway time for a small reactor with negligible cooling, exothermic reaction, and heat capacity $1000\,\text{J/K}$, heat-generation rate $100\,\text{W}$ at room temperature, activation energy $80\,\text{kJ/mol}$.
\end{exercise}

\begin{exercise}
Estimate the consequence of a lithium-ion battery cell thermal runaway in a multi-cell pack with no thermal isolation.
\end{exercise}

\begin{exercise}
Estimate the wind speed at which a tall slender stack would experience VIV lock-in with its first mode (assume cylinder diameter $1\,\text{m}$, structural frequency $2\,\text{Hz}$, Strouhal $0.2$).
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate the buckling of a slender column with imperfections; identify the load-displacement behaviour and the imperfection sensitivity.
\end{exercise}

\begin{exercise}
Simulate the response of a single-DOF resonant structure to forced oscillation at and near resonance; identify the bandwidth and the Q.
\end{exercise}

\begin{exercise}
Simulate thermal runaway in a reactor with stochastic disturbance; observe the trajectory from steady state through runaway.
\end{exercise}

\begin{exercise}
Simulate the rounding-error accumulation in a long-duration time integration; identify the precision requirement for acceptable accuracy.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a thin-walled aerospace structure for axial compression with adequate buckling margin and reasonable mass.
\end{exercise}

\begin{exercise}
Design a thermal-runaway protection system for a polymerisation reactor: emergency cooling, depressurisation, inhibitor injection.
\end{exercise}

\begin{exercise}
Design a vibration-isolation system for a precision instrument exposed to ambient vibration at $10$ to $100\,\text{Hz}$.
\end{exercise}

\begin{exercise}
Design a flutter-clearance test program for a new aircraft wing.
\end{exercise}

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A team's structure has buckled in service at $80\%$ of predicted critical load. State three engineering hypotheses.
\end{exercise}

\begin{exercise}
A team's chemical reactor is being uprated for higher throughput; the original safety analysis assumed lower throughput. State the engineering response and the role of process-hazard analysis.
\end{exercise}

\begin{exercise}
A team's tall building shows wind-induced vibration that exceeds occupant comfort criteria. State the engineering options including tuned mass damper, structural stiffness modification, and damping enhancement.
\end{exercise}

\begin{exercise}
A team is investigating a software-induced control failure; the system was working until a specific input sequence triggered the failure. The same investigation also finds that a structural analysis used a textbook column-end fixity assumption that does not match the actual installation. State the engineering response addressing both: state-space safety analysis for the software side, and bracketing the actual end fixity for the structural side.
\end{exercise}
