\chapter{Transport networks}
\label{vol12:ch04}

\epigraph{Cities and countries have the transport systems they have because of decisions taken decades ago by people who did not consult anyone now living.}{Author, after Mumford}

\chapmeta{Half-life: medium-life. Archetypes: transport, scaling, ethics. Exercise target: 25. Project track: simulation (transport-corridor redesign). Hazard class: medium (chronic safety; multi-decade lock-in). Reader path default: standard, with the modal-mix and decarbonisation sections core, the urban and inter-city sections standard, and the maritime and aviation sections mastery. Named cases: Boeing 737 MAX MCAS (\cite{acc:knkt-lion-air-610-2019}; \cite{acc:house-737max-2020}); Genoa Morandi Bridge collapse 2018 (\cite{acc:mit-morandi-2018}).}

\noindent
Transport is the engineering of motion at civilization scale. Approximately $25\%$ of global energy use goes to moving people and goods; roughly $20\%$ of global CO$_2$ emissions; the world's largest single category of engineered infrastructure (paved roads alone: more than $30 \times 10^6\,\text{km}$); and a substantial fraction of the global engineering workforce. The transport system the engineering profession inherits is the result of more than a century of investment in road and rail networks, of post-WWII car-centric urbanism, of the containerisation revolution of the 1960s, and of the aviation expansion since 1970. The transition the chapter describes is to a transport system that delivers comparable accessibility at a substantially lower carbon, land, and safety cost.

\section{Modal mix: road, rail, sea, air}\pathtag{core}

Global passenger transport in 2023 totalled approximately $50 \times 10^{12}\,\text{passenger-km/yr}$ (passenger-kilometres), of which roughly $75\%$ by road (private car dominant), $10\%$ by rail, $10\%$ by air, $5\%$ other (bus, coach, two-wheeler in developing world). Global freight transport in 2023 totalled approximately $130 \times 10^{12}\,\text{tonne-km/yr}$ (tonne-kilometres), of which roughly $30\%$ by road, $20\%$ by rail, $40\%$ by sea (dominant in international freight), $5\%$ by inland waterway, $5\%$ by air (high-value low-mass). The modal mix differs by region: rail-and-bus heavy in Europe and Japan, road heavy in the USA and Australia, sea heavy in international trade everywhere.

Each mode has a characteristic energy intensity (MJ/passenger-km or MJ/tonne-km), carbon intensity (gCO$_2$/passenger-km), and infrastructure intensity (capital per unit transport service). The cross-modal comparison is the foundation of decarbonisation analysis. As a representative range: passenger road (private car): $\approx 2\,\text{MJ/p-km}$; passenger rail: $\approx 0.5\,\text{MJ/p-km}$; aviation: $\approx 1.5$-$2\,\text{MJ/p-km}$. Freight: road truck: $\approx 2\,\text{MJ/t-km}$; rail: $\approx 0.3\,\text{MJ/t-km}$; sea: $\approx 0.1\,\text{MJ/t-km}$. The order-of-magnitude differences across modes are the central engineering fact.

\begin{archetype}[Scaling: modal energy intensity]
Sea freight is roughly $20\times$ more energy-efficient per tonne-km than road freight; rail freight is roughly $7\times$ more efficient. The modal shifts that decarbonisation pathways depend on are thus high-leverage: moving $10\%$ of road freight to rail produces emissions reductions equivalent to $70\%$ of the rail sector's existing emissions. The constraint is not energy intensity but accessibility, time-to-delivery, and the infrastructure to support modal shift (intermodal terminals, rail capacity, port-hinterland connections). The engineering task is to make modal shift feasible at the operational level.
\end{archetype}

\section{Urban transport networks}\pathtag{core}

Urban transport is the dominant share of most cities' passenger transport. The system the engineering profession inherits is post-war car-centric in most US cities and many Asia-Pacific cities; a hybrid of mass transit, cycling, walking, and car in most European and historic-Asian cities. The car-centric urbanism the chapter's failure section discusses produced characteristic outcomes: low residential density, high per-capita VMT (vehicle-miles-travelled), high per-capita carbon, high road-traffic injury and fatality rates, low transit ridership, and the disinvestment from non-car infrastructure that reinforces the car-dependence. The reversal that contemporary urban-transport engineering is engaged in: high-frequency mass transit; protected cycling networks; pedestrian-prioritised streetscapes; demand-pricing of road space; transit-oriented development.

The mode-share data is illustrative. Greater London's mode share for passenger trips: $\approx 35\%$ rail and bus, $\approx 30\%$ walking, $\approx 25\%$ car, $\approx 5\%$ cycle (post-2020 expansion). Houston Texas: $\approx 90\%$ car, $\approx 5\%$ bus, $\approx 5\%$ other. The two cities have similar populations and similar per-capita GDP; the difference in mode share reflects $70$ years of differing infrastructure investment and zoning policy. The engineering implication: city-scale mode share is the cumulative result of decades of decisions; restructuring it requires decades of sustained intervention.

\begin{principle}[Induced demand]
Road capacity expansion in congested urban regions typically does not reduce congestion; it produces additional vehicle trips (induced demand) that re-saturate the new capacity within $5$-$10$ years \cite{paper:duranton-turner-2011}. The empirical elasticity of VMT with respect to lane-miles in congested regions is approximately $1.0$: a $10\%$ expansion of lane-miles produces approximately $10\%$ more VMT. The engineering implication: capacity expansion is not the appropriate response to congestion in mature urban regions; demand management (pricing, modal shift, land-use intensification) is. The failure mode of ignoring this principle is the ``one more lane'' urban-highway expansion that has consumed several trillion dollars of US capital over $70$ years without producing the promised congestion relief.
\end{principle}

\begin{estimation}
\textbf{Question.} Estimate the urban land area consumed by automobile infrastructure (roads, parking, fuel and service) in a typical US city of 1 million people.

\textbf{Reasoning.} Per-capita road length: $\approx 3$ m of street per person ($3{,}000$ km of street for 1 million people, at $\approx 10\,\text{m}$ width including sidewalks gives $30\,\text{km}^2$). Parking: $\approx 4$ spaces per car (home, work, shopping, miscellaneous), at $30\,\text{m}^2$ per space including circulation, with $\approx 0.8$ cars per person, gives $\approx 100\,\text{m}^2$ per person or $100\,\text{km}^2$ for 1 million people. Total: $\approx 130\,\text{km}^2$, against a typical US urban land area of $\approx 500\,\text{km}^2$ for a city of this size. Conclusion: automobile infrastructure consumes approximately $25\%$-$30\%$ of urban land in typical US cities, more than residential housing by some accounts. The land-use implication is the central engineering-economics insight about car-centric urbanism.
\end{estimation}

\section{Inter-city and freight networks}\pathtag{standard}

Inter-city passenger transport in the developed world is split between road (long-distance car and bus), rail (intercity), and air (over $500\,\text{km}$ typical break-even). The choice depends on distance, frequency, terminal access, and the relative cost-and-time profile. The mode-choice models that traffic engineers use (logit choice models with travel time, cost, and convenience attributes) predict the modal shares with reasonable accuracy. The policy choices that change the modal shares are the infrastructure investments that change the cost-and-time profiles of each mode.

Freight networks combine truck (dominant in short-haul, $<500$ km), rail (dominant in long-haul bulk and intermodal), barge (dominant in heavy-bulk where waterways exist), and sea (dominant in international). The economics is well-defined: each mode has a fixed-cost component (terminal handling, transfers) and a variable-cost component (per-km cost); the modal choice for a given shipment is the combination that minimises total cost given the time and reliability constraints. The intermodal terminal (rail-to-truck transfer; ship-to-truck-or-rail transfer at port) is the engineering object where the cost-trade-offs are realised; the global trend toward larger ships and larger terminals continues to lower the per-tonne-km cost of sea freight.

\begin{archetype}[Transport: hub-and-spoke vs point-to-point]
Freight and passenger networks are organised as hub-and-spoke or point-to-point. Hub-and-spoke (airline networks; intermodal rail; postal sorting) consolidates traffic at hubs to achieve high load factors on the trunk routes at the cost of additional handling and transit time. Point-to-point (truck freight; high-frequency bus; some emerging airline networks) trades load factor for directness. The engineering choice depends on the demand density, the time sensitivity of the cargo or passengers, and the cost structure of consolidation. Most large networks are hybrids: hub-and-spoke for the trunk, point-to-point or local distribution for the last-mile.
\end{archetype}

\section{Maritime and shipping}\pathtag{mastery}

Maritime shipping carries approximately $80\%$ of global trade by volume and $70\%$ by value. The container shipping industry, since the 1960s containerisation revolution, has reduced per-tonne international transport costs by an order of magnitude and is one of the major engineering enablers of globalisation. The contemporary fleet: approximately $100{,}000$ commercial vessels; the largest container ships approach $24{,}000$ TEU (twenty-foot equivalent units); the largest oil tankers approach $400{,}000$ DWT; the largest bulk carriers approach $400{,}000$ DWT. The industry's emissions are roughly $3\%$ of global CO$_2$ (\cite{std:imo-ghg-2020}) and the decarbonisation challenge is among the hardest of any sector: large engines that have run on heavy fuel oil for a century, with no obvious drop-in low-carbon replacement at the scale required.

The current decarbonisation pathway combines efficiency improvements (slow steaming; bulbous bows; air lubrication; sail-assist propulsion), alternative fuels (LNG as transitional; ammonia, methanol, and hydrogen as longer-term), and operational changes (port-side electrification of berthed ships; just-in-time arrival to reduce port congestion). The IMO 2023 strategy targets net-zero by 2050; the engineering trajectory to achieve this is not yet settled, with the alternative-fuel choice the central uncertainty.

\section{Aviation}\pathtag{mastery}

Commercial aviation carries approximately $4 \times 10^9$ passengers/yr and $60$ million tonnes of cargo/yr, producing approximately $2.5\%$ of global CO$_2$ emissions (and a comparable additional radiative-forcing contribution from contrails and high-altitude NO$_x$). The fleet is roughly $25{,}000$ commercial aircraft; the industry is highly concentrated (Boeing and Airbus duopoly for narrow- and wide-body; Embraer and the now-defunct CSeries-now-A220 for regional). The decarbonisation challenge is the hardest in transport: aircraft cannot use batteries at scale because of weight; alternative fuels (sustainable aviation fuel from biomass or synthesised from hydrogen and captured CO$_2$) are technically feasible but currently a small fraction of supply and substantially more expensive than fossil jet fuel.

The Boeing 737 MAX MCAS accidents (\cite{acc:knkt-lion-air-610-2019}; \cite{acc:house-737max-2020}) discussed in Volume X provide the contemporary case for the chapter's transport-safety analysis. The accidents and their aftermath illustrate the failure of the certification regime to keep up with derivative-aircraft modifications that change the handling characteristics in ways that the previous certification basis did not contemplate. The institutional response (the FAA Aircraft Certification Reform Act of 2020; the international airworthiness re-certification of the 737 MAX) is the regulatory adjustment that the failure forced.

\section{Decarbonising transport}\pathtag{core}

The transport-decarbonisation pathway combines four moves: (1) modal shift from high-carbon to low-carbon modes (cars to rail; trucks to rail; air to high-speed rail for sub-$1{,}000$-km trips); (2) electrification of road vehicles (light-duty cars; bus and short-haul freight; eventually long-haul trucks); (3) alternative fuels for the modes that resist electrification (sustainable aviation fuel; ammonia or methanol or hydrogen for shipping); (4) demand management (urban-form changes to reduce trip lengths; remote-work to reduce trip frequency; mode-pricing to internalise the externalities). Each move is at a different stage of deployment: electric light-duty vehicles are at $\approx 20\%$ of new sales globally in 2024 and rising rapidly; electric trucks are emerging; alternative aviation fuel is single-digit percent; modal shift is region-specific.

The Genoa Morandi Bridge collapse \cite{acc:mit-morandi-2018} of 2018, discussed in Volume X Chapter 14, illustrates the maintenance dimension of transport infrastructure: the existing road network is itself ageing and requires sustained renewal investment that has been deferred in many developed-world countries for decades. The decarbonisation transition coincides with a maintenance transition; the engineering profession is asked to deliver both.

\section{Failure: car-centric urbanism and its costs}\pathtag{core}

Car-centric urbanism is the failure mode the chapter ends with. The pattern is well-documented: post-WWII US suburban development pattern (zoning that separated land uses; minimum-parking requirements that subsidised driving; highway construction that displaced urban neighbourhoods; transit disinvestment); replication of the pattern in Australia and parts of Canada; partial replication in Asia-Pacific cities (especially Southeast Asia). The consequences: roughly $40{,}000$ US road-traffic fatalities per year (more than $1{,}000{,}000$ globally per year per WHO; \cite{web:who-road-safety-2023}); per-capita transport CO$_2$ emissions in car-centric cities roughly $5\times$ the comparable European city; chronic time costs of congestion measured in hundreds of hours per worker per year; the displacement of pedestrian, cycling, and child-independent-mobility from urban space; and the spatial inequality produced by the access-by-car presumption.

The engineering profession's contribution to this failure mode: the traffic-engineering tradition of the mid-20th century that took car throughput as the optimisation objective and the road system as the unit of analysis. The contribution to the reversal: the contemporary multimodal-transport-engineering tradition (associated with the Institute of Transportation Engineers' \emph{Designing Walkable Urban Thoroughfares} and the National Association of City Transportation Officials' \emph{Urban Street Design Guide}) that takes accessibility as the optimisation objective and the streetscape as the unit of analysis. The engineering content of the reversal is well-developed; the political-economic execution is the binding constraint.

\begin{principle}[Streets are public-space allocation problems]
A street is a finite right-of-way that must be allocated among uses: through-vehicle movement, parking, pedestrian movement, cycle movement, public-transit movement, drainage, utilities, frontage activity. The 20th-century engineering tradition allocated most of the right-of-way to through-vehicle and parking. The 21st-century engineering tradition is renegotiating the allocation under different priorities: safety, equity, climate, public health. The engineering analysis is the cross-modal capacity, safety, and accessibility trade-off at the street scale; the institutional analysis is the political process by which the renegotiation is conducted.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Transport-corridor redesign]
Choose a transport corridor in your region (an arterial road; an inter-city rail line; a freight corridor; an airport access link). Document the current modal mix, the current capacity and demand, the current safety record, and the current emissions. Identify the principal design problems against a 21st-century set of priorities (carbon, safety, accessibility, equity, climate-resilience). Design a redesign that addresses the problems; specify the engineering interventions, the institutional changes required, the estimated capital and operational cost, the distributive consequences. Document the just-transition strategy for the affected stakeholders (commuters, freight operators, adjacent residents, transit-dependent populations).
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A passenger car carrying $1.5$ passengers (typical occupancy) and using $7\,\text{L}/100\,\text{km}$ of gasoline produces approximately $2.3\,\text{kg CO}_2/\text{L}$ of combustion emissions. Compute the gCO$_2$/p-km. Compare to an electric car at $20\,\text{kWh}/100\,\text{km}$ on a grid at $300\,\text{gCO}_2/\text{kWh}$ and on a grid at $50\,\text{gCO}_2/\text{kWh}$.
\end{exercise}

\begin{exercise}
A high-speed rail train at $300\,\text{km/h}$ uses $20\,\text{kWh/seat-km} \times 10^{-3}$ (i.e., $20\,\text{Wh/seat-km}$) with $80\%$ seat occupancy. Compute the gCO$_2$/p-km on a low-carbon grid at $50\,\text{gCO}_2/\text{kWh}$ and compare to short-haul aviation at $150\,\text{gCO}_2/\text{p-km}$.
\end{exercise}

\begin{exercise}
A container ship carrying $20{,}000$ TEU at $20$ knots over a $20{,}000\,\text{km}$ Asia-to-Europe route uses approximately $200\,\text{tonnes}$ of bunker fuel/day at $3.1\,\text{tCO}_2/\text{tonne}$. Compute the gCO$_2$/t-km for a $10$-tonne container (typical loaded mass).
\end{exercise}

\begin{exercise}
A US urban highway expansion adds $4$ lane-miles to a $20$-lane-mile corridor. Using an induced-demand elasticity of $0.8$, compute the predicted VMT increase. Compare to the original capacity expansion percentage and identify the congestion-relief implication.
\end{exercise}

\begin{exercise}
A city of $1$ million people targets a $50\%$ shift from car trips to transit and active modes over $20$ years. Compute the required transit capacity expansion, assuming current transit share of $10\%$ and average car trip length of $10\,\text{km}$.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the modal-energy-intensity ranking from first principles (rolling resistance, aerodynamic drag, gradient resistance, propulsive efficiency, vehicle mass per passenger). Identify the dominant terms for each mode.
\end{exercise}

\begin{exercise}
Derive the optimal hub-and-spoke network topology for a given demand matrix, assuming a per-hub fixed cost and a per-leg variable cost. Identify the regime in which point-to-point service is preferred.
\end{exercise}

\begin{exercise}
Derive the relationship between transit-line frequency, station spacing, and door-to-door travel time for a typical urban-transit user.
\end{exercise}

\begin{exercise}
Derive the carbon-payback time for an electric vehicle relative to its internal-combustion counterpart, given manufacturing emissions for the battery and the operational emissions difference.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the lithium-ion battery demand for full electrification of the global light-duty vehicle fleet ($\approx 1.5 \times 10^9$ cars). Compare to current battery production capacity.
\end{exercise}

\begin{exercise}
Estimate the rail freight capacity expansion required to absorb a $30\%$ modal shift from long-haul truck freight in a major economy. Identify the infrastructure investments required.
\end{exercise}

\begin{exercise}
Estimate the parking-land-area reduction available from $50\%$ shift from private cars to shared autonomous vehicles in a typical US city.
\end{exercise}

\begin{exercise}
Estimate the sustainable aviation fuel production required to decarbonise the global commercial aviation fleet, and the bioresource or hydrogen-and-CO$_2$ input required to produce it.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Implement a multi-modal transport-network simulation for a city with given OD (origin-destination) demand. Sweep the road-pricing and transit-investment parameters and identify the configurations that produce target mode-share shifts.
\end{exercise}

\begin{exercise}
Simulate the EV-charging-network capacity for a region with given EV adoption trajectory. Identify the grid-investment implications and the optimal mix of fast and slow chargers.
\end{exercise}

\begin{exercise}
Simulate the freight network optimisation under different intermodal-cost assumptions. Identify the modal-split sensitivity to terminal-handling cost and to fuel-price changes.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a bus rapid transit (BRT) corridor for a city of $500{,}000$ people. Specify the route, the station spacing, the vehicle and infrastructure capacity, the integration with other modes, the operational frequency, the cost and travel-time savings.
\end{exercise}

\begin{exercise}
Design a complete-streets retrofit for a $5\,\text{km}$ urban arterial currently configured for car throughput. Specify the lane allocation, the pedestrian and cycling facilities, the transit priority, the parking management, the safety design.
\end{exercise}

\begin{exercise}
Design a hydrogen-fueling network for long-haul truck freight on a major freight corridor. Specify the station spacing, the hydrogen supply (electrolysis, transport, storage), the cost-recovery model.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: the US transport-emissions problem cannot be solved by vehicle electrification alone; modal shift and urban-form change are required. Argue the opposite. Identify the empirical evidence; identify the political-economic constraints.
\end{exercise}

\begin{exercise}
Discuss the Boeing 737 MAX MCAS accidents as a transport-safety and certification-regime failure. Identify the engineering and regulatory decisions; identify the implications for the certification of derivative aircraft.
\end{exercise}

\begin{exercise}
Argue: the engineering profession has a responsibility to engage with land-use planning because transport and land-use are inseparable. Argue the opposite. Identify the implications for engineering-education curricula and for professional-society activity.
\end{exercise}

\begin{exercise}
Discuss the just-transition implications of transport decarbonisation (auto-industry workforce; fuel-distribution workforce; transit-dependent low-income populations). Identify the engineering and institutional interventions.
\end{exercise}

\begin{exercise}
Discuss the trade-off between autonomous-vehicle deployment and transit investment as competing decarbonisation strategies. Identify the empirical uncertainties and the engineering judgments.
\end{exercise}

\begin{exercise}
Discuss the Genoa Morandi Bridge collapse as a transport-infrastructure maintenance failure. Identify the engineering decisions and the institutional context; identify the implications for the global stock of ageing transport infrastructure.
\end{exercise}
