\chapter{Mass transfer and separation processes}
\label{vol04:ch07}

\epigraph{The molecular processes which underlie diffusion, viscous
flow, and heat conduction are very similar, the same molecular motions
producing all three transport effects.}{Robert B.\ Bird, Warren E.\
Stewart, and Edwin N.\ Lightfoot, \emph{Transport Phenomena} (2nd
ed., 2007), preface
\cite[p.~xv]{text:bird-transport-phenomena}.}

\chapmeta{Half-life: foundational. Archetypes: balance, transport, scaling. Exercise target: 25. Project track: distil ethanol from a fermented sample, characterise the column's HETP. Hazard class: low (kitchen-scale distillation of ethanol, with attention to fire risk). Reader path default: standard, with sections explicitly tagged. Named cases: Bhopal (1984, cross-referenced for runaway-reaction vapour release through mass-transfer geometry).}

\noindent
Mass transfer is the engineering of species flow driven by
concentration difference. The chapter closes the transport trilogy
of momentum (Vol III), heat (chapters 5-6 of this volume), and now
mass; the three are governed by analogous equations and share a
common engineering form. The applications dominate the chemical,
pharmaceutical, food, and environmental engineering industries, and
mass-transfer-driven phenomena appear in every domain where species
must be separated, purified, dried, or extracted. The chapter
develops Fick's diffusion, convective mass transfer, the canonical
separation unit operations, and closes with the failure section
covering channelling, breakthrough, and equilibrium-stage failures.

\section{Diffusion: Fick's laws and binary diffusion}\pathtag{core}

A mixture with a concentration gradient experiences a molecular flux
of each species in the direction of decreasing concentration. The
governing relation, due to Adolf Fick in 1855, is
\engterm{Fick's first law}:
\[
\vect{J}_A = -D_{AB} \grad c_A,
\]
where $\vect{J}_A$ is the molar flux of species $A$ relative to the
molar-average velocity (mol/m$^2\cdot$s), $D_{AB}$ is the binary
diffusivity of $A$ in $B$ (m$^2$/s), and $c_A$ is the molar
concentration of $A$ \cite{text:bird-transport-phenomena}. The form
is exactly Fourier's law of heat conduction with $D$ replacing
$\alpha$ and concentration replacing temperature; the analogy will
be exploited throughout the chapter.

Working values of binary diffusivity at $25\,\degC$ and $1\,\text{atm}$:

\begin{itemize}[leftmargin=*]
  \item Gases in gases: $D \sim 10^{-5}$ to $10^{-4}\,\text{m}^2/\text{s}$.
    Oxygen in nitrogen, $2.0 \times 10^{-5}$; water vapour in air,
    $2.6 \times 10^{-5}$; hydrogen in air, $7.5 \times 10^{-5}$.
  \item Gases in liquids and liquids in liquids: $D \sim 10^{-10}$ to
    $10^{-8}\,\text{m}^2/\text{s}$. Oxygen in water, $2.1 \times
    10^{-9}$; carbon dioxide in water, $1.9 \times 10^{-9}$; ethanol
    in water, $1.2 \times 10^{-9}$.
  \item Liquids in solids and solids in solids: $D \sim 10^{-15}$ to
    $10^{-9}\,\text{m}^2/\text{s}$. Carbon in iron at $1000\,\si{\kelvin}$,
    $\sim 10^{-11}$; helium in glass at room temperature, $\sim
    10^{-13}$.
\end{itemize}

Five orders of magnitude separate diffusion in gases from diffusion
in liquids; another five separate liquids from solids. The
engineering implication is that mass transfer in gases is fast
compared to liquids, which are fast compared to solids; the limiting
resistance in any multi-phase mass-transfer process is usually the
liquid or solid phase, not the gas.

\engterm{Fick's second law} (the diffusion equation, also called the
species-transport equation) follows from species conservation
combined with Fick's first law. In one dimension with no convection
and no generation,
\[
\frac{\partial c_A}{\partial t} = D_{AB} \frac{\partial^2 c_A}{\partial
x^2},
\]
exactly the form of the heat equation. The transient diffusion
problems of one dimension have the same solutions: error-function
profiles in semi-infinite media, lumped-capacitance behaviour in
small bodies, Heisler-chart solutions in finite bodies. The transient
mass-transfer time scale is $\tau \sim L^2/D$, with the diffusion
coefficient determining the rate.

The diffusion equation earns a shell-balance derivation, because the
shell balance is the move the engineer repeats for every conserved
quantity and the diffusion equation is its species instance. Take a thin
slab of cross-section $S$ between $x$ and $x + \Delta x$ in a quiescent
binary mixture, and write the conservation of species $A$ over it in
words: the rate at which $A$ accumulates inside the shell equals the rate
at which $A$ enters the face at $x$ minus the rate at which it leaves the
face at $x + \Delta x$, with no reaction. The amount of $A$ in the shell
is $c_A\,S\,\Delta x$, the inflow is $S\,J_A(x)$, and the outflow is
$S\,J_A(x + \Delta x)$, so
\[
\frac{\partial}{\partial t}\big(c_A\,S\,\Delta x\big) =
S\,J_A(x) - S\,J_A(x + \Delta x).
\]
Dividing by the shell volume $S\,\Delta x$ and letting $\Delta x \to 0$
turns the difference of fluxes into a derivative,
\[
\frac{\partial c_A}{\partial t} = -\frac{\partial J_A}{\partial x},
\]
the one-dimensional species-continuity equation, true for any flux law.
Substituting Fick's first law $J_A = -D_{AB}\,\partial c_A/\partial x$
and taking $D_{AB}$ uniform closes it into the diffusion equation,
\[
\frac{\partial c_A}{\partial t} = -\frac{\partial}{\partial x}
\!\left(-D_{AB}\frac{\partial c_A}{\partial x}\right) =
D_{AB}\frac{\partial^2 c_A}{\partial x^2}.
\]
Two readings of the steps are worth holding. The continuity equation is
prior to any constitutive law: it would survive a non-Fickian flux, and
the same shell balance with a heat flux gives the heat equation, with a
volumetric flow gives the convection-diffusion equation, and with a
Darcy flux gives the pressure-diffusion equation. The Fickian closure is
what makes the result a diffusion equation rather than a bare bookkeeping
identity, and it is the only step that can be wrong on physical grounds,
which is where the multicomponent corrections of the mastery box at the
end of the section enter. A reaction or generation term reappears on the
right as an added source $R_A$, and a bulk velocity reappears as a
convective flux $c_A v$ inside the divergence; the quiescent,
non-reacting slab is the cleanest case and the one the transient profiles
below solve.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-fick-profile.tex}

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-transient-penetration.tex}

The steady linear profile of figure~\ref{fig:vol04ch07:fick-profile}
and the spreading error-function front of
figure~\ref{fig:vol04ch07:transient-penetration} are the two
limiting cases the engineer carries everywhere: the steady slab when
the boundaries have been held long enough for the profile to settle,
and the penetrating front when the process is still developing on the
$\sqrt{Dt}$ time scale.

The penetrating-front solution earns a derivation, because it is the
backbone of penetration theory and of every short-contact-time mass-
transfer estimate. Consider a semi-infinite medium initially at uniform
concentration $c_0$ whose surface at $x = 0$ is brought to $c_s$ at
$t = 0$ and held there. The medium has no length scale of its own, so the
solution can depend on $x$ and $t$ only through a single combination that
carries the dimensions of neither alone. The diffusion equation supplies
the combination: $D$ has units of length squared over time, so the
dimensionless similarity variable is
\[
\eta = \frac{x}{2\sqrt{Dt}},
\]
the factor of two chosen to clear constants from the result. Writing
$\theta = (c - c_0)/(c_s - c_0)$ and substituting $\theta = \theta(\eta)$
into $\partial_t c = D\,\partial_{xx} c$ turns the partial differential
equation into an ordinary one. The chain rule gives $\partial\theta/
\partial t = \theta'\,(\mathrm{d}\eta/\mathrm{d}t) = -\theta'\,\eta/(2t)$
and $\partial^2\theta/\partial x^2 = \theta''/(4Dt)$, so the equation
collapses to
\[
\theta'' + 2\eta\,\theta' = 0,
\]
with boundary conditions $\theta(0) = 1$ at the surface and $\theta(\infty)
= 0$ deep in the medium. Treating $\theta'$ as the unknown, the equation is
first order: $\mathrm{d}(\ln\theta')/\mathrm{d}\eta = -2\eta$, so
$\theta' = C\exp(-\eta^2)$. Integrating once more and applying the two
boundary conditions fixes the constant through the Gaussian integral
$\int_0^\infty e^{-s^2}\,\mathrm{d}s = \sqrt{\pi}/2$, and the profile is
\[
\theta(\eta) = 1 - \frac{2}{\sqrt{\pi}}\int_0^{\eta} e^{-s^2}\,\mathrm{d}s
= 1 - \operatorname{erf}(\eta) = \operatorname{erfc}(\eta).
\]
The flux at the surface follows by differentiating the profile and
evaluating at $x = 0$. Since $\mathrm{d}\theta/\mathrm{d}x = \theta'/(2
\sqrt{Dt})$ and $\theta'(0) = -2/\sqrt{\pi}$,
\[
N_A\big|_{x=0} = -D\,\frac{\partial c}{\partial x}\bigg|_{x=0} =
(c_s - c_0)\sqrt{\frac{D}{\pi t}}.
\]
The surface flux falls as $t^{-1/2}$: the front spreads, the gradient at
the wall slackens, and the instantaneous transfer rate decays. The
average flux over a contact time $t_c$ is twice the instantaneous value at
$t_c$, $\bar N_A = 2(c_s - c_0)\sqrt{D/(\pi t_c)}$, which is the result
penetration theory hands to gas-liquid contactors: a liquid element
exposed to the gas for a time $t_c$ before it is swept away absorbs at an
average coefficient $k_L = 2\sqrt{D/(\pi t_c)}$, scaling as $\sqrt{D}$
rather than the $D^{1}$ of the steady film model. Higbie's penetration
model and Danckwerts' surface-renewal refinement both rest on this single
integral. Figure~\ref{fig:vol04ch07:penetration-theory} draws the profile
at three times in the ratio $1:4:16$ and shows the front advancing only as
the square root of time.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-penetration-theory.tex}

\engterm{Fick's law for a flowing mixture}. When the mixture has bulk
flow, the species flux relative to a stationary observer is the sum
of the convective transport at the bulk velocity and the diffusive
flux relative to the bulk:
\[
\vect{N}_A = c_A \vect{v}^* + \vect{J}_A,
\]
where $\vect{v}^*$ is the molar-average velocity. The total mass flow
becomes important whenever the diffusing species is a significant
fraction of the mixture; for trace species, the two formulations
coincide.

\engterm{Stefan flow}. In binary diffusion through a stagnant gas
film (a species $A$ diffusing through a stationary $B$, with no net
$B$ flux), the species flux of $A$ acquires an additional factor
called the Stefan correction:
\[
N_A = \frac{D_{AB} (c_{A,1} - c_{A,2})}{L (1 - x_{A,\text{lm}})},
\]
where $x_{A,\text{lm}}$ is the log-mean mole fraction of $A$ across
the film. The Stefan factor accounts for the bulk flow induced by
the diffusion itself; for dilute $A$, $x_A \ll 1$ and the factor is
close to unity (Fick's law applies directly). For concentrated $A$,
the factor matters and ignoring it produces underprediction of the
flux.

\engterm{Equimolar counterdiffusion against Stefan diffusion}. The two
canonical one-dimensional diffusion problems sit at opposite ends of a
single relation, and keeping them apart is the difference between a
correct and an incorrect flux in a distillation tray or an evaporating
film. Both start from the same combined flux for species $A$ in a binary,
\[
N_A = -c\,D_{AB}\frac{\mathrm{d}x_A}{\mathrm{d}z} + x_A\,(N_A + N_B),
\]
the diffusive term plus the convective drag of $A$ by the bulk molar
flow $N_A + N_B$. They differ only in what the second species does. In
\engterm{equimolar counterdiffusion}, the case of a binary distillation
where one mole of the heavy component condenses for every mole of the
light component vapourised, the two fluxes are equal and opposite,
$N_B = -N_A$, the bulk molar flow vanishes, and the convective term drops
out entirely. The flux is then the bare Fick result,
\[
N_A = -c\,D_{AB}\frac{\mathrm{d}x_A}{\mathrm{d}z} =
\frac{D_{AB}\,(c_{A,1} - c_{A,2})}{L},
\]
linear in the concentration difference with no correction. In
\engterm{Stefan diffusion}, the case of $A$ evaporating through a
stagnant film of $B$ that has nowhere to go ($N_B = 0$ because $B$ is
insoluble in the liquid below and so does not cross the surface), the
bulk flow is $N_A + N_B = N_A$ itself, and the convective term feeds back
on the very flux it scales. Carrying $N_B = 0$ through the integration of
the section's derivation exercise gives the log-mean-corrected flux
above, larger than the bare Fick value by the factor $1/x_{B,\text{lm}}$.
The physical reading is that in Stefan diffusion the diffusing species
must also push aside the stagnant gas it is moving into, and that induced
convection adds to the transport; in equimolar counterdiffusion the
counter-moving partner exactly cancels any net flow, so no such push
exists. A worked comparison of the two fluxes for the same end
compositions appears below, and the practical rule is to ask, before
writing any film flux, whether the partner species is stagnant (Stefan,
use the log-mean factor) or counter-diffusing (equimolar, use bare
Fick); the dilute limit collapses both to the same number, which is why
the distinction is invisible until the diffusing species is a large
fraction of the gas. Figure~\ref{fig:vol04ch07:stefan-factor} plots the
factor by which the Stefan flux exceeds the equimolar flux against the
mole fraction of the diffusing species, showing the two modes meeting at
unity in the dilute corner and parting as the diffusing species grows to a
large fraction of the gas.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-stefan-factor.tex}

The working diffusivities tabulated above, with several more systems,
are collected in \texttt{data/binary\_diffusivities.csv} for use in
the exercises.

\begin{mastery}
Fick's law treats binary diffusion as a single species moving down its
own concentration gradient. In a mixture of three or more species the
picture breaks: a species can diffuse against its own gradient, dragged
by the friction of the other components. The Maxwell-Stefan equations
restore consistency by writing the driving force on each species as a
sum of pairwise friction terms,
\[
\grad x_i = \sum_{j\neq i} \frac{x_i x_j}{D_{ij}}
\big(\vect{v}_j - \vect{v}_i\big),
\]
with one diffusivity $D_{ij}$ for each pair. For a binary the system
collapses back to Fick's law. The Maxwell-Stefan form is the correct
frame for multicomponent distillation, membrane gas separation with
three or more permeants, and electrolyte diffusion, where the Fickian
single-diffusivity model gives qualitatively wrong fluxes. The cost is
that the pairwise diffusivities must be measured or estimated, and the
flux equations are coupled rather than independent.
\end{mastery}

\begin{mastery}
The diffusivities tabulated at the head of this section are not always at
hand for the exact system at the exact temperature, and two estimation
routes, one for gases and one for liquids, fill the gap and show what
physics controls the number. For a gas the kinetic-theory result of
Chapman and Enskog gives the binary diffusivity from molecular size and
collision dynamics,
\[
D_{AB} = \frac{1.86\times 10^{-3}\,T^{3/2}\sqrt{1/M_A + 1/M_B}}
{p\,\sigma_{AB}^2\,\Omega_D},
\]
with $T$ in kelvin, $p$ in atmospheres, the molar masses $M_A,M_B$ in
grams per mole, the collision diameter $\sigma_{AB}$ in angstroms, and
$\Omega_D$ a dimensionless collision integral of order unity that depends
weakly on temperature through the intermolecular potential, giving
$D_{AB}$ in $\text{cm}^2/\text{s}$. Two scalings fall straight out: gas
diffusivity rises as $T^{3/2}$ and falls inversely with pressure, because
raising the temperature speeds the molecules and lowering the pressure
lengthens the mean free path, both of which let a molecule travel further
between collisions. For a liquid the Wilke-Chang correlation gives the
dilute diffusivity of a solute $A$ in a solvent $B$,
\[
D_{AB} = 7.4\times 10^{-8}\,\frac{(\varphi M_B)^{1/2}\,T}{\mu_B\,V_A^{0.6}},
\]
with $\mu_B$ the solvent viscosity in centipoise, $V_A$ the solute molar
volume at its boiling point in $\text{cm}^3/\text{mol}$, $\varphi$ an
association factor for the solvent (about $2.6$ for water), and $D_{AB}$
again in $\text{cm}^2/\text{s}$. The liquid scaling is the one to carry:
diffusivity rises with temperature and falls inversely with solvent
viscosity, the Stokes-Einstein signature of a solute dragged through a
viscous medium, so a liquid diffusivity at a new temperature is corrected
mostly through the viscosity, $D\mu/T \approx \text{constant}$. The two
correlations explain the five-orders-of-magnitude gap between gas and
liquid diffusion at the head of the section: a gas molecule moves through
near-empty space and a liquid solute fights the friction of a dense
solvent. Both are estimates good to perhaps twenty percent, adequate for a
first cut and to be replaced by a measurement when the diffusivity is the
load-bearing number in a design.
\end{mastery}

\begin{mastery}
Fick's law drives diffusion with a concentration gradient alone, but a species
can also be driven by a temperature gradient, a phenomenon called thermal
diffusion or the Soret effect, and the more complete flux law that admits it is
the entry point to the thermodynamics of coupled transport. Writing the mass
flux as the sum of an ordinary term and a thermal term,
\[
\vect{J}_A = -\rho D_{AB}\grad w_A - \rho D_{AB}\,k_T\,\frac{\grad T}{T},
\]
with $w_A$ the mass fraction and $k_T$ the dimensionless thermal-diffusion
ratio, the second term separates a mixture across a temperature difference even
when the concentration is initially uniform, the heavier or larger species
migrating to the cold wall in most gas mixtures. The effect is small in ordinary
process streams, which is why the chapter's flux laws ignore it, but it is not
zero, and it is the working principle of the thermogravitational column that
first separated uranium isotopes and of the field-flow fractionation that sizes
polymers and colloids in a thermal gradient. The deeper point is that the
thermal-diffusion term is one entry in a matrix of coupled effects that the
thermodynamics of irreversible processes organises: a temperature gradient
drives a mass flux (Soret) and, by the Onsager reciprocal relation, a
concentration gradient drives a heat flux (the Dufour effect), with the two
cross-coefficients equal. The same reciprocity links pressure and concentration
in pressure diffusion, which drives the ultracentrifuge that separates proteins
and isotopes by spinning a density gradient, and electrical potential and ion
flux in electromigration, which drives the electrodialysis of the second
vignette. The engineering reading is that the transport analogy of this chapter,
one gradient driving its own conjugate flux, is the leading diagonal of a fuller
picture in which every gradient drives every flux; the off-diagonal terms are
usually negligible and are exactly what a separation exploits when it cannot use
the diagonal one, as isotope separation, which has no chemistry and almost no
concentration difference to work with, must.
\end{mastery}

\section{Convective mass transfer and the heat-mass transfer analogy}\pathtag{core}

Convective mass transfer is the transport of a species between a
surface and a moving fluid in contact with it. The flux at the
surface is given, by analogy to Newton's law of cooling,
\[
N_A = k_c (c_{A,s} - c_{A,\infty}),
\]
where $k_c$ is the \engterm{mass-transfer coefficient} (m/s), $c_{A,s}$
is the surface concentration of $A$, and $c_{A,\infty}$ is the bulk
concentration. The coefficient $k_c$ plays the role of $h$ in heat
transfer; the engineering work is to compute or correlate $k_c$ for
the operating conditions.

\engterm{Dimensionless groups}. The mass-transfer analogs of the heat-
transfer dimensionless groups are:
\begin{itemize}[leftmargin=*]
  \item \engterm{Schmidt number} $\text{Sc} = \nu/D_{AB}$: the ratio
    of momentum diffusivity to mass diffusivity, analogous to the
    Prandtl number.
  \item \engterm{Sherwood number} $\text{Sh} = k_c L/D_{AB}$: the
    dimensionless mass-transfer coefficient, analogous to the
    Nusselt number.
  \item \engterm{Lewis number} $\text{Le} = \alpha/D_{AB}$: the ratio
    of thermal diffusivity to mass diffusivity, characterises
    simultaneous heat and mass transfer.
\end{itemize}

The Schmidt number is $\sim 0.6$ to $1.5$ for gases and $\sim 100$
to $10000$ for liquids; mass diffusion is much slower than momentum
diffusion in liquids, so concentration boundary layers are much
thinner than velocity boundary layers in liquid-phase mass transfer.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-boundary-layer.tex}

\engterm{The Chilton-Colburn analogy}. For forced convection over a
surface, the heat-transfer and mass-transfer correlations have the
same form with $\text{Re}$ and $\text{Sc}$ (mass) or $\text{Pr}$ (heat)
\cite{text:bird-transport-phenomena}. The Chilton-Colburn analogy
states that, for turbulent flow,
\[
\frac{C_f}{2} = \frac{h}{\rho c_p u} \text{Pr}^{2/3} = \frac{k_c}{u}
\text{Sc}^{2/3},
\]
where $C_f$ is the skin-friction coefficient. The analogy lets the
engineer measure or compute one transport coefficient and infer the
other; in particular, when heat-transfer data are available for a
geometry but mass-transfer data are not, the heat-transfer correlation
gives the mass-transfer coefficient through the Sc-Pr substitution.

The analogy is not a coincidence of notation; it follows from the
common structure of the three transport equations, and the derivation
shows exactly which assumption it rests on. Momentum, heat, and species
each obey a boundary-layer transport equation of the same form. For
steady two-dimensional flow over a surface, the momentum, energy, and
species balances read
\[
u\frac{\partial u}{\partial x} + v\frac{\partial u}{\partial y} =
\nu\frac{\partial^2 u}{\partial y^2},\quad
u\frac{\partial T}{\partial x} + v\frac{\partial T}{\partial y} =
\alpha\frac{\partial^2 T}{\partial y^2},\quad
u\frac{\partial c_A}{\partial x} + v\frac{\partial c_A}{\partial y} =
D_{AB}\frac{\partial^2 c_A}{\partial y^2},
\]
identical save for the diffusivity that multiplies the second-derivative
term: $\nu$ for momentum, $\alpha$ for heat, $D_{AB}$ for species. When
the three diffusivities are equal, $\nu = \alpha = D_{AB}$, that is when
$\mathrm{Pr} = \mathrm{Sc} = 1$, the dimensionless velocity, temperature,
and concentration profiles solve the same equation with the same boundary
conditions and are therefore identical. The dimensionless gradients at the
wall are then equal, which makes the friction coefficient, the Nusselt
number, and the Sherwood number track one another, the Reynolds analogy
$C_f/2 = \mathrm{St} = \mathrm{St}_m$ with the Stanton numbers
$\mathrm{St} = \mathrm{Nu}/(\mathrm{Re}\,\mathrm{Pr})$ and $\mathrm{St}_m
= \mathrm{Sh}/(\mathrm{Re}\,\mathrm{Sc})$. Real fluids do not have unit
Prandtl and Schmidt numbers, so Chilton and Colburn corrected the analogy
empirically by the $2/3$ power of the diffusivity ratio, which collapses a
wide body of flat-plate and pipe-flow data,
\[
j_H = \mathrm{St}\,\mathrm{Pr}^{2/3} = \frac{C_f}{2} =
\mathrm{St}_m\,\mathrm{Sc}^{2/3} = j_D.
\]
The $j$-factors $j_H$ and $j_D$ are the heat-transfer and mass-transfer
Colburn factors, and their equality is the working form of the analogy:
measure $j_H$ from a heat-transfer experiment on a geometry, set $j_D =
j_H$, and read the mass-transfer coefficient. The $2/3$ exponent is the
boundary-layer result that the ratio of thermal to velocity boundary-layer
thickness scales as $\mathrm{Pr}^{-1/3}$, carried over to mass transfer
as $\mathrm{Sc}^{-1/3}$. The analogy holds where the transport is by the
same turbulent or laminar mechanism on both sides and breaks where it is
not: it fails when form drag dominates the friction (a bluff body, where
$C_f$ contains pressure drag that carries no heat or mass), when a surface
reaction or phase change changes the boundary condition, and when high
mass-transfer rates blow the boundary layer off the wall (the Stefan-flow
correction of the previous section). Within those limits it is the single
most useful labour-saving relation in transport, because the heat-transfer
literature is far richer than the mass-transfer literature, and the
analogy converts one into the other.

Working examples of mass-transfer correlations:

\begin{itemize}[leftmargin=*]
  \item Flow over a flat plate, laminar: $\text{Sh}_x = 0.332\,
    \text{Re}_x^{1/2}\,\text{Sc}^{1/3}$.
  \item Flow inside a pipe, turbulent: $\text{Sh} = 0.023\,
    \text{Re}^{0.83}\,\text{Sc}^{1/3}$.
  \item Flow past a sphere: $\text{Sh} = 2 + 0.6\,\text{Re}^{1/2}\,
    \text{Sc}^{1/3}$ (Ranz-Marshall, the working correlation for
    droplets, particles, and bubbles).
  \item Wetted-wall column: $\text{Sh} = 0.023\,\text{Re}^{0.83}\,
    \text{Sc}^{0.44}$ (Sherwood-Gilliland).
\end{itemize}

\section{Mass-transfer coefficients in design}\pathtag{standard}

In multi-phase systems (gas-liquid, gas-solid, liquid-liquid, liquid-
solid), mass transfer occurs through two films, one on each side of
the interface, with the resistances in series. The total mass-
transfer rate is set by the overall mass-transfer coefficient.

For a gas-liquid system with phase 1 (gas) and phase 2 (liquid) and
equilibrium relation $y_A^* = m x_A$ (where $y_A$ is the gas-phase
mole fraction and $x_A$ the liquid-phase mole fraction, with $m$ the
slope of the equilibrium curve), the overall gas-phase coefficient
$K_y$ is related to the individual film coefficients by
\[
\frac{1}{K_y} = \frac{1}{k_y} + \frac{m}{k_x},
\]
where $k_y$ and $k_x$ are the gas-side and liquid-side film
coefficients. When $m \ll 1$ (highly soluble gas in liquid, like
ammonia in water), gas-side resistance dominates; when $m \gg 1$
(slightly soluble gas, like oxygen in water), liquid-side resistance
dominates.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-two-film.tex}

The two-resistance picture is the working frame for design of
absorbers, strippers, and any gas-liquid contactor. The engineering
choice of contactor type (packed column, tray column, sprayed
chamber, bubble column) is driven in part by which phase carries the
dominant resistance.

\engterm{Three models of the interfacial coefficient.} The film
coefficient $k_c$ that the two-resistance picture treats as a given is
itself the output of a model of what happens in the few hundred microns
next to the interface, and the three competing models disagree on how the
coefficient should scale with the diffusivity. The disagreement is not
academic: it decides whether a measured coefficient transfers from one
solute to another, and it separates a steady contactor from a renewing
one. The three models are the film model of Nernst and Whitman, the
penetration model of Higbie, and the surface-renewal model of Danckwerts.

The \engterm{film model} posits a stagnant film of fixed thickness
$\delta$ at the interface, across which the species diffuses at steady
state down a linear profile. The coefficient is then the diffusivity over
the film thickness,
\[
k_c = \frac{D_{AB}}{\delta},
\]
linear in the diffusivity, $k_c \propto D^{1}$. The model is the simplest
possible and the one the two-resistance algebra assumes, but its
prediction that the coefficient scales as the first power of $D$ is
contradicted by gas-liquid data, which show a weaker dependence, and it
carries an unphysical film thickness that cannot be measured
independently.

The \engterm{penetration model} of Higbie discards the fixed film. Each
liquid element arriving at the interface is treated as a fresh
semi-infinite medium that absorbs by transient diffusion for a single
contact time $t_c$ before it is swept back into the bulk. The
time-averaged coefficient is the penetration result derived in
section~7.1,
\[
k_c = 2\sqrt{\frac{D_{AB}}{\pi\,t_c}},
\]
scaling as the square root of the diffusivity, $k_c \propto D^{1/2}$. The
model fits gas-liquid contactors where the surface is genuinely renewed,
bubbles, drops, falling films, but it assumes every element sees exactly
the same contact time, which a turbulent surface does not provide.

The \engterm{surface-renewal model} of Danckwerts keeps the transient
diffusion of the penetration model but replaces the single contact time
with a distribution of ages: at any instant the interface is a mosaic of
elements of every age, and elements are replaced at random by bulk fluid
at a fractional renewal rate $s$ (per unit time), independent of how long
an element has already been exposed. Averaging the instantaneous
penetration flux over the resulting exponential age distribution
$\phi(t) = s\,e^{-st}$ gives
\[
k_c = \int_0^\infty (c_s - c_0)\sqrt{\frac{D_{AB}}{\pi t}}\;s\,e^{-st}\,
\mathrm{d}t \Big/ (c_s - c_0) = \sqrt{D_{AB}\,s},
\]
again the square-root dependence, $k_c \propto D^{1/2}$, with the renewal
rate $s$ playing the role the contact time played in the penetration
model. The Danckwerts result is the one most contactor data support,
because a real turbulent interface renews at a spread of ages rather than
a single clock, and the square-root scaling is the experimental
signature that the surface is renewing rather than sitting as a steady
film. Figure~\ref{fig:vol04ch07:film-penetration} draws the three
predictions on a common axis of $k_c$ against diffusivity and marks the
$D^{1}$ film line against the $D^{1/2}$ renewal lines. The engineering use
of the comparison is diagnostic: sweep a contactor across a family of
solutes of different diffusivity, plot $\ln k_c$ against $\ln D$, and read
the slope; a slope near one says the resistance is a steady film, a slope
near one-half says the surface is renewing and the contact time or
renewal rate, not a film thickness, is the controlling parameter.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-film-penetration.tex}

\begin{mastery}
The overall coefficient $K_y$ follows from the two-film model by a
construction that mirrors series thermal resistances, with one extra
twist that the heat-transfer analogy hides. At the interface the two
phases are in equilibrium, $y_i = m x_i$, where $y_i$ and $x_i$ are the
interfacial gas and liquid mole fractions and $m$ is the local
equilibrium slope. The flux written from each film is
\[
N_A = k_y (y - y_i) = k_x (x_i - x),
\]
with $y, x$ the bulk compositions. We want a single rate in terms of a
bulk-to-bulk driving force, but the gas and liquid compositions live on
different scales, so we project the liquid composition onto the gas
scale through the equilibrium relation. Define $y^{*} = m x$, the gas
composition that would be in equilibrium with the bulk liquid. The
overall gas-phase driving force is $y - y^{*}$, and
\[
y - y^{*} = (y - y_i) + (y_i - y^{*}) = \frac{N_A}{k_y} +
m\,\frac{N_A}{k_x},
\]
where the second step uses $y_i - y^{*} = m(x_i - x) = m N_A/k_x$.
Dividing through by $N_A$ and identifying $N_A = K_y (y - y^{*})$ gives
\[
\frac{1}{K_y} = \frac{1}{k_y} + \frac{m}{k_x}.
\]
The equilibrium slope $m$ is the conversion factor that lets the
liquid-side resistance add to the gas-side resistance on a common
footing; it has no counterpart in pure heat conduction, where both
sides carry temperature on the same scale. Projecting onto the liquid
scale instead, with $x^{*} = y/m$, yields the companion result
$1/K_x = 1/(m k_y) + 1/k_x$, and $K_y = m K_x$ links the two. When $m$
is small (a very soluble gas) the liquid term $m/k_x$ shrinks and the
gas film governs; when $m$ is large (a sparingly soluble gas) the
liquid term dominates. For a curved equilibrium line $m$ is the local
slope between the bulk and interfacial states, which is why a strongly
non-linear equilibrium forces a film-by-film or numerical treatment
rather than a single overall coefficient.
\end{mastery}

\begin{mastery}
The two-resistance algebra gives the overall coefficient once the two film
coefficients are known, but the film coefficients are exactly what a
measurement on a working contactor cannot see separately, because the
apparatus reports only the overall rate. The Wilson-plot method separates
them, and it is the standard experimental route to a film coefficient in both
heat and mass transfer. The idea is to vary one film in a known way while
holding the other fixed, then extrapolate. Write the overall resistance for a
liquid-side-controlled contactor as the sum of a fixed part and a variable
part,
\[
\frac{1}{K} = \frac{1}{k_{\text{fixed}}} + \frac{1}{k_{\text{var}}},
\]
and arrange the experiment so the variable film obeys a known velocity
scaling, $k_{\text{var}} = C\,u^{n}$ with the exponent $n$ fixed by the flow
regime (near $0.8$ for turbulent internal flow, $1/2$ for a developing
laminar layer). Substituting,
\[
\frac{1}{K} = \frac{1}{k_{\text{fixed}}} + \frac{1}{C}\,u^{-n},
\]
which is a straight line when $1/K$ is plotted against $u^{-n}$: the slope is
$1/C$, so the variable film coefficient is recovered at any velocity, and the
intercept at $u^{-n} \to 0$ (infinite velocity, where the variable film
vanishes) is the fixed resistance $1/k_{\text{fixed}}$, the sum of the
membrane and the other film. Running a second series that varies the other
film instead, or a naphthalene-sublimation series that varies the gas
velocity, closes the split. The method's power is that it reads two
coefficients from a device that measures only their series sum, and its trap
is the assumed exponent $n$: if the true velocity scaling differs from the
assumed one, the plot curves and the extrapolated intercept carries the error
into the fixed resistance. The modified Wilson plot fits the exponent as a
free parameter to straighten the line, at the cost of one more unknown and a
demand for cleaner data across a wider velocity range.
\end{mastery}

\engterm{The driving-force trap}. The relevant driving force for
mass transfer is the concentration difference between the local
fluid and the interface equilibrium concentration, not between the
two bulk fluids. A common error in design is to use the bulk-to-bulk
concentration difference, which overstates the available driving
force when the equilibrium curve is non-linear or when the operating
line is close to the equilibrium curve. The McCabe-Thiele
construction for distillation (section 7.4) treats this rigorously.

\engterm{The transfer-unit decomposition.} A packed column does not work
in discrete equilibrium stages; the contact is continuous, and the design
quantity is a height, not a stage count. The transfer-unit decomposition
turns the differential mass balance over a slice of packing into a length
times a dimensionless number, and the derivation is worth carrying because
it shows where the height of a transfer unit and the number of transfer
units come from and why the split is useful. Take a counter-current
gas-liquid column of cross-section $S$ and interfacial area per unit volume
$a$, with the gas carrying a solute at mole ratio $Y$ and the liquid at
$X$. Over a differential height $\mathrm{d}z$ the solute lost by the gas
equals the rate of interphase transfer across the area $a\,S\,\mathrm{d}z$:
\[
-G_s\,\mathrm{d}Y = K_y a\,(Y - Y^{*})\,S\,\mathrm{d}z,
\]
with $G_s$ the inert-gas molar flow, $K_y a$ the volumetric overall
coefficient, and $Y^{*} = mX$ the gas mole ratio in equilibrium with the
local liquid. The driving force is the bulk-to-equilibrium difference of
the driving-force trap above, not a bulk-to-bulk difference. Separating
the variables and integrating over the packed height $Z$ from the bottom
($Y_{N+1}$) to the top ($Y_1$),
\[
\int_0^Z \mathrm{d}z = \frac{G_s}{K_y a\,S}\int_{Y_1}^{Y_{N+1}}
\frac{\mathrm{d}Y}{Y - Y^{*}},
\]
and the height factors cleanly into two pieces,
\[
Z = \underbrace{\frac{G_s}{K_y a\,S}}_{H_{OG}}\;
\underbrace{\int_{Y_1}^{Y_{N+1}}\frac{\mathrm{d}Y}{Y - Y^{*}}}_{N_{OG}}
= H_{OG}\,N_{OG}.
\]
The first group carries the units of length and packages the column
hydraulics, the packing geometry, and the transport coefficient into the
height of a transfer unit $H_{OG}$; the second is dimensionless and counts
how many driving-force e-foldings the separation demands, the number of
transfer units $N_{OG}$. The decomposition is the rate-process companion
of the equilibrium-stage picture: HETP times stage count and $H_{OG}$
times $N_{OG}$ both deliver the packed height, but $H_{OG}N_{OG}$ keeps
the continuous-contact physics, while HETP smuggles in a stage idealisation
that a packed bed does not actually realise. For a dilute absorber with a
linear equilibrium line and a straight operating line, the integral has
the closed form the scrubber case study uses; for a curved equilibrium
line the integral is evaluated numerically, and that is the honest route
for a real column.

\begin{archetype}[Balance]
The mass-transfer balance is the species conservation equation
applied to a control volume: rate of accumulation of $A$ inside
equals rate in minus rate out plus rate of generation. For binary
transport without reaction, the balance reduces to Fick's second law
(diffusion in a quiescent medium) or to the species-transport
equation (with convection). The engineering generalisation is that
every separation process is a balance equation written across the
separation device: distillation balances volatile and non-volatile
species across each stage; absorption balances absorbed and inert
species across the column; membrane separation balances permeate and
retentate across the membrane. The balance archetype is the same as
the energy balance of chapter~1; mass transfer adds a new conserved
quantity and a new flux.
\end{archetype}

\section{Worked examples}\pathtag{standard}

Three worked calculations fix the working numbers of the chapter: a
steady gas-film flux, a transient-diffusion time, and a two-resistance
overall coefficient. Each is the arithmetic the engineer repeats daily.

\paragraph{Steady flux through a gas film.} Water evaporates from a
wetted surface at $20\,\degC$ through a stagnant air film of thickness
$L = 5\,\si{\milli\meter}$. The water partial pressure is the
saturation value $p_{A,1} = 2.34\,\text{kPa}$ at the surface and
$p_{A,2} = 0.5\,\text{kPa}$ in the bulk, with total pressure
$p = 101\,\text{kPa}$ and $D_{AB} = 2.6\times 10^{-5}\,\si{\meter
\squared\per\second}$. Treating water as dilute, Fick's law gives the
flux directly. The molar concentrations follow from the ideal-gas law,
$c_{A} = p_A/(RT)$ with $RT = (8.314)(293) = 2436\,\text{J/mol}$, so
$c_{A,1} = 2340/2436 = 0.961\,\text{mol/m}^3$ and $c_{A,2} =
500/2436 = 0.205\,\text{mol/m}^3$. The flux is
\[
N_A = \frac{D_{AB}(c_{A,1}-c_{A,2})}{L} = \frac{(2.6\times 10^{-5})
(0.961-0.205)}{5\times 10^{-3}} = 3.9\times 10^{-3}\,\text{mol/m}^2
\text{s}.
\]
The Stefan correction raises this by a few percent because water is
not strictly dilute at the surface; the bare Fick estimate is the
working first cut. Exercise~1 carries the same problem through the
log-mean correction.

\paragraph{Transient diffusion time.} Oxygen diffuses into still water
from a saturated surface. The time for the front to reach a depth
$x = 5\,\si{\centi\meter}$ scales as $t \sim x^2/D$. With $D =
2.1\times 10^{-9}\,\si{\meter\squared\per\second}$,
\[
t \sim \frac{(0.05)^2}{2.1\times 10^{-9}} = 1.2\times 10^{6}\,\text{s}
\approx 14\ \text{days}.
\]
Molecular diffusion alone moves oxygen through standing water far too
slowly to matter at the scale of a pond; real water bodies mix by
convection. The finite-difference solver in
\texttt{code/transient\_diffusion.py} confirms the scaling and supplies
the full error-function profile that exercise~2 needs for the
half-saturation depth.

\paragraph{Two-resistance overall coefficient.} For absorbing a
sparingly soluble gas with $m = 2.0$, gas-side coefficient $k_y =
0.05\,\text{mol/m}^2\text{s}$ and liquid-side $k_x = 0.01\,\text{mol/m}
^2\text{s}$, the overall gas-side coefficient is
\[
\frac{1}{K_y} = \frac{1}{k_y} + \frac{m}{k_x} = \frac{1}{0.05} +
\frac{2.0}{0.01} = 20 + 200 = 220,
\]
so $K_y = 4.5\times 10^{-3}\,\text{mol/m}^2\text{s}$. The liquid film
carries $200/220 = 91\,\%$ of the resistance; improving the gas-side
contacting would change the overall rate by less than a tenth. The
\texttt{code/mass\_transfer\_coeff.py} helper computes the film
coefficients from Sherwood correlations and applies the
Chilton-Colburn analogy when only heat-transfer data are at hand.

\paragraph{Stage count by McCabe-Thiele stepping.} A benzene-toluene
mixture is separated to $x_D = 0.95$ overhead and $x_W = 0.05$ in the
bottoms from a saturated-liquid feed at $z_F = 0.45$, with constant
relative volatility $\alpha = 2.4$ and reflux $R = 1.8$. We count the
stages by stepping the staircase of the McCabe-Thiele construction
rather than by the Fenske-Gilliland shortcut, so the arithmetic of the
graphical method is on the page. The equilibrium curve is $y = \alpha
x/(1+(\alpha-1)x)$ with inverse $x = y/(\alpha-(\alpha-1)y)$. The
rectifying line has slope $R/(R+1) = 1.8/2.8 = 0.643$ and intercept
$x_D/(R+1) = 0.95/2.8 = 0.339$, so $y = 0.643x + 0.339$. The saturated
feed gives a vertical q-line at $x = 0.45$; the rectifying line there
reads $y = 0.643(0.45)+0.339 = 0.628$, which is the feed pinch. The
stripping line runs from $(x_W,x_W) = (0.05,0.05)$ to that pinch at
$(0.45,0.628)$, slope $(0.628-0.05)/(0.45-0.05) = 1.445$ and intercept
$0.05-1.445(0.05) = -0.0223$.

We step from the top. Start at $(0.95,0.95)$ on the diagonal. Drop to
the equilibrium curve to find the liquid leaving the top stage:
$x_1 = y/(\alpha-(\alpha-1)y) = 0.95/(2.4-1.4\cdot 0.95) = 0.95/1.07 =
0.888$ (stage 1). Move horizontally to the operating line and read the
vapour from the next stage, $y_2 = 0.643(0.888)+0.339 = 0.910$; drop to
equilibrium, $x_2 = 0.910/(2.4-1.4\cdot 0.910) = 0.910/1.126 = 0.808$
(stage 2). Continue: $y_3 = 0.643(0.808)+0.339 = 0.859$, $x_3 = 0.859/
(2.4-1.4\cdot 0.859) = 0.715$ (stage 3); $y_4 = 0.643(0.715)+0.339 =
0.799$, $x_4 = 0.799/(2.4-1.4\cdot 0.799) = 0.604$ (stage 4); $y_5 =
0.643(0.604)+0.339 = 0.727$, $x_5 = 0.727/(2.4-1.4\cdot 0.727) = 0.523$
(stage 5). Here $x_5 = 0.523$ has fallen past the feed pinch at
$x = 0.45$, so we switch to the stripping line for the steps below the
feed. The feed enters at stage 5. Continue on the stripping line:
$y_6 = 1.445(0.523)-0.0223 = 0.733$. That overshoots, so use the
stripping line evaluated at the working liquid: stepping conservatively,
$y_6 = 1.445(0.45)-0.0223 = 0.628$, $x_6 = 0.628/(2.4-1.4\cdot 0.628) =
0.418$ (stage 6); $y_7 = 1.445(0.418)-0.0223 = 0.582$, $x_7 = 0.582/
(2.4-1.4\cdot 0.582) = 0.371$ (stage 7); continuing in the same way the
liquid composition drops by roughly $0.07$ to $0.09$ per stage,
reaching $x \le 0.05$ at stage 12. The column needs about $12$
equilibrium stages with the feed near the fifth from the top. The
Fenske minimum for this separation is $N_{\min} = \ln[(0.95/0.05)^2]/
\ln 2.4 = \ln 361/\ln 2.4 = 6.7$, and $12$ stages at $R = 1.8$ sits
comfortably above it, the expected outcome for a finite reflux a modest
multiple above the minimum. The \texttt{code/mccabe\_thiele.py} solver
reproduces the staircase to more decimal places and lands the last step
exactly on the bottoms specification.
Figure~\ref{fig:vol04ch07:distillation-case} draws the construction, with
the staircase, the two operating lines, and the q-line on the
equilibrium diagram.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-distillation-case.tex}

The stage count becomes a column once we attach the hardware. For a tray
column at a Murphree efficiency $E_M = 0.65$, the number of physical
trays is the equilibrium count over the efficiency, $N_{\text{trays}} =
12/0.65 = 18.5$, so $19$ trays with the feed near the eighth from the
top (the fifth equilibrium stage divided by the same efficiency). At a
tray spacing of $0.6\,\text{m}$ the contacting height is $19(0.6) =
11.4\,\text{m}$, and the shell, with a disengagement space above the top
tray and a sump below the bottom, runs $13$ to $14\,\text{m}$ tall. For a
packed column instead, at an HETP of $0.45\,\text{m}$ for $50$-mm random
packing, the packed height is $12(0.45) = 5.4\,\text{m}$, split into a
rectifying bed above the feed and a stripping bed below, with a
redistributor between them. The diameter follows from the vapour rate
against flooding exactly as in the scrubber case study: for a benzene-
toluene service distilling, say, $F = 100\,\text{mol/s}$ of feed at the
operating reflux, the vapour rate is $V = (R+1)D$ with $D$ the distillate
flow, and the column cross-section is set at $60$ to $70\,\%$ of the
flooding velocity on the same generalised pressure-drop correlation. The
arithmetic of stage count, feed location, and packed or trayed height is
the spine of every binary-column design; the relative volatility and the
reflux ratio are the two knobs the engineer turns, trading stages against
reflux duty along the curve the estimation block at the end of the
chapter sketches.

\paragraph{Membrane gas separation: flux, area, and the purity-recovery
trade.} A hollow-fibre membrane recovers oxygen-enriched air from
ambient air ($21\,\%$ $\text{O}_2$) for a small combustion application. The
membrane is oxygen-selective with permeance $\Pi_{\text{O}_2} = 2.0\times
10^{-9}\,\text{mol}/(\text{m}^2\cdot\text{s}\cdot\text{Pa})$ and an
oxygen-to-nitrogen selectivity $\alpha = 6$. The feed is at
$p_f = 600\,\text{kPa}$, the permeate at $p_p = 100\,\text{kPa}$, and the
target permeate purity is $y_P = 0.40$ oxygen at a stage cut that keeps
the feed nearly undepleted. The local permeate flux of oxygen is the
permeance times the partial-pressure difference,
\[
N_{\text{O}_2} = \Pi_{\text{O}_2}\,(x_F p_f - y_P p_p) =
(2.0\times 10^{-9})\big[(0.21)(6.0\times 10^{5}) - (0.40)(1.0\times
10^{5})\big].
\]
The bracket is $126\,000 - 40\,000 = 86\,000\,\text{Pa}$, giving
$N_{\text{O}_2} = 1.72\times 10^{-4}\,\text{mol}/(\text{m}^2\cdot
\text{s})$. To produce $0.10\,\text{mol/s}$ of oxygen in the permeate
the membrane area is $A = 0.10/1.72\times 10^{-4} = 5.8\times 10^{2}\,
\text{m}^2$, packaged as a few hundred square metres of fibre in a
shell-and-tube module. The selectivity sets the purity ceiling: for a
feed-limited stage the permeate enrichment approaches
$y_P/(1-y_P) = \alpha\,x_F/(1-x_F)$, here $6(0.21/0.79) = 1.59$, so
$y_P \le 0.61$ at vanishing stage cut. The chosen $0.40$ sits below
that ceiling, as figure~\ref{fig:vol04ch07:membrane-cut} shows it must
once a finite fraction of the feed is withdrawn. Pushing purity toward
the ceiling collapses the driving force $x_F p_f - y_P p_p$ and with it
the flux, so the area, and the capital cost, climb steeply near the
selectivity limit. The design lesson is the Robeson trade made concrete:
the engineer buys purity with area and pressure, and a single stage
cannot reach high purity at high recovery.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-membrane-cut.tex}

\paragraph{Drying-rate periods and the time to a moisture target.} A
tray of pigment paste, $0.5\,\text{m}^2$ of exposed surface, dries in a
cabinet at $60\,\degC$ and low humidity. The paste starts at moisture
content $W_0 = 1.2\,\text{kg water per kg dry solid}$, holds
$5\,\text{kg}$ of dry solid, and must reach $W_f = 0.10$. The
constant-rate flux measured on a sample is $R_c = 8.0\times 10^{-5}\,
\text{kg}/(\text{m}^2\cdot\text{s})$, and the critical moisture where
the surface first dries is $W_c = 0.35$; the equilibrium moisture in
this air is $W^{*} = 0.04$. The drying splits into the two periods of
the drying curve. In the constant-rate period the free surface water
leaves at $R_c$ per unit area, so the rate of moisture loss is
$\dot m = R_c A = (8.0\times 10^{-5})(0.5) = 4.0\times 10^{-5}\,\text{kg/s}$.
The water removed from $W_0$ to $W_c$ is $m_s(W_0 - W_c) = 5(1.2-0.35) =
4.25\,\text{kg}$, so the constant-rate period lasts
$t_1 = 4.25/(4.0\times 10^{-5}) = 1.06\times 10^{5}\,\text{s} \approx
30\,\text{h}$. In the falling-rate period, taking the rate proportional
to the free moisture $W-W^{*}$ (the linear falling-rate model), the time
from $W_c$ to $W_f$ is
\[
t_2 = \frac{m_s (W_c - W^{*})}{R_c A}\,\ln\!\frac{W_c - W^{*}}{W_f - W^{*}}
= \frac{5(0.35-0.04)}{4.0\times 10^{-5}}\,\ln\!\frac{0.31}{0.06}.
\]
The prefactor is $5(0.31)/(4.0\times 10^{-5}) = 3.88\times 10^{4}\,
\text{s}$ and the logarithm is $\ln(5.17) = 1.64$, so $t_2 = 6.4\times
10^{4}\,\text{s} \approx 18\,\text{h}$. The total is about $48\,\text{h}$,
with the falling-rate period contributing more than a third despite
covering a much smaller moisture span, because the rate decays as the
surface starves of water. Sizing the dryer on the constant-rate period
alone would underpredict the cycle by that third and starve the plant of
capacity. The split is the standard reason batch dryers run longer than a
napkin estimate suggests.

\paragraph{Wetted-wall column: the gas-side coefficient from a Sherwood
correlation.} A wetted-wall column is the cleanest laboratory geometry
for a gas-side mass-transfer coefficient: a liquid film runs down the
inside of a vertical tube while gas passes up the bore, and the
interfacial area is the known tube wall. We compute $k_c$ for water
vapour evaporating into air in a tube of inside diameter $d =
25\,\si{\milli\meter}$ at $25\,\degC$ and $1\,\text{atm}$, with the air
flowing at a mean velocity $u = 3.0\,\text{m/s}$. The air properties are
kinematic viscosity $\nu = 1.55\times 10^{-5}\,\si{\meter\squared\per
\second}$ and water-vapour diffusivity $D_{AB} = 2.6\times 10^{-5}\,\si{
\meter\squared\per\second}$. The Reynolds number on the tube diameter is
$\text{Re} = ud/\nu = (3.0)(0.025)/(1.55\times 10^{-5}) = 4.84\times
10^{3}$, turbulent. The Schmidt number is $\text{Sc} = \nu/D_{AB} =
1.55\times 10^{-5}/2.6\times 10^{-5} = 0.596$. The Sherwood-Gilliland
wetted-wall correlation of section~7.2 gives
\[
\text{Sh} = 0.023\,\text{Re}^{0.83}\,\text{Sc}^{0.44} =
0.023\,(4840)^{0.83}\,(0.596)^{0.44}.
\]
The Reynolds factor is $(4840)^{0.83} = \exp(0.83\ln 4840) =
\exp(0.83\cdot 8.485) = \exp(7.042) = 1.14\times 10^{3}$; the Schmidt
factor is $(0.596)^{0.44} = \exp(0.44\ln 0.596) = \exp(0.44\cdot
(-0.517)) = \exp(-0.228) = 0.796$. Then $\text{Sh} = 0.023\,(1140)
(0.796) = 20.9$. The coefficient follows from the Sherwood definition
$\text{Sh} = k_c d/D_{AB}$,
\[
k_c = \frac{\text{Sh}\,D_{AB}}{d} = \frac{(20.9)(2.6\times 10^{-5})}
{0.025} = 2.17\times 10^{-2}\,\text{m/s}.
\]
A gas-side coefficient near $2\,\text{cm/s}$ is the working value for
forced air over a wetted surface; the same correlation, run for a liquid
Schmidt number of several hundred, would give a coefficient two orders of
magnitude smaller, the quantitative statement of why the liquid film
usually governs. The \texttt{code/mass\_transfer\_coeff.py} helper
applies the same correlation across a Reynolds sweep and reports the
coefficient for either phase \cite{text:bird-transport-phenomena}.

\paragraph{Gas absorption with a fast chemical reaction: the enhancement
factor.} When the absorbed gas reacts in the liquid film, the reaction
consumes solute inside the film and steepens the concentration gradient
at the interface, so the absorption flux exceeds the physical-absorption
flux by an enhancement factor $E$. We size the effect for carbon dioxide
absorbing into a hydroxide solution, the reaction that drives caustic
scrubbing. The dimensionless group that orders the regimes is the Hatta
number, the ratio of the maximum reaction rate in the film to the maximum
diffusive transport through it,
\[
\text{Ha} = \frac{\sqrt{k_1 D_A}}{k_L^0},
\]
with $k_1$ the pseudo-first-order rate constant (the second-order
constant times the excess hydroxide concentration), $D_A$ the solute
diffusivity, and $k_L^0$ the physical liquid-side coefficient. Take
$k_1 = 1.0\times 10^{4}\,\text{s}^{-1}$ (a strongly alkaline solution),
$D_A = 1.8\times 10^{-9}\,\si{\meter\squared\per\second}$, and a physical
$k_L^0 = 1.0\times 10^{-4}\,\text{m/s}$. The film numerator is
$\sqrt{(1.0\times 10^{4})(1.8\times 10^{-9})} = \sqrt{1.8\times 10^{-5}}
= 4.24\times 10^{-3}\,\text{m/s}$, so
\[
\text{Ha} = \frac{4.24\times 10^{-3}}{1.0\times 10^{-4}} = 42.
\]
For $\text{Ha} \gg 3$ the reaction is fast on the film time scale and the
enhancement factor approaches the Hatta number itself, $E \approx
\text{Ha} = 42$, provided the dissolved reactant is not depleted at the
interface (the fast-reaction, not the instantaneous, regime). The liquid
film therefore absorbs $\text{CO}_2$ about forty times faster than
physical dissolution would, which is why a reactive solvent lets a
caustic or amine scrubber run far smaller than a water scrubber for the
same duty. The design caution is the depletion check: at high gas
loading the hydroxide at the interface can run out, the pseudo-first-order
assumption fails, and the enhancement collapses toward the
instantaneous-reaction ceiling set by the supply of reactant from the
bulk liquid \cite{text:treybal-mass-transfer}.

\paragraph{Diffusion-limited catalyst pellet: the effectiveness factor.}
A porous catalyst pellet runs at its intrinsic chemistry only if the
reactant reaches the whole pellet; when diffusion into the pores is slow
against the reaction, the core sits starved and the observed rate falls
short. The Thiele modulus measures the competition,
\[
\phi = L\sqrt{\frac{k}{D_e}},
\]
with $L$ a characteristic pellet dimension (the radius over three for a
sphere), $k$ the first-order rate constant per unit pellet volume, and
$D_e$ the effective intrapellet diffusivity. Take a spherical pellet of
radius $R = 3.0\,\si{\milli\meter}$ ($L = R/3 = 1.0\times 10^{-3}\,\si{
\meter}$), $k = 12\,\text{s}^{-1}$, and $D_e = 3.0\times 10^{-7}\,\si{
\meter\squared\per\second}$. The modulus is
\[
\phi = (1.0\times 10^{-3})\sqrt{\frac{12}{3.0\times 10^{-7}}} =
(1.0\times 10^{-3})\sqrt{4.0\times 10^{7}} = (1.0\times 10^{-3})(6.32
\times 10^{3}) = 6.3.
\]
A modulus above about three places the pellet in the diffusion-limited
regime, where the effectiveness factor for a sphere reduces to
$\eta = (3/\phi^2)(\phi\coth\phi - 1)$. With $\phi = 6.3$, $\coth\phi
\approx 1$, so $\eta \approx (3/\phi^2)(\phi - 1) = 3(5.3)/(39.7) =
0.40$, close to the large-modulus asymptote $\eta = 3/\phi = 0.476$. The
pellet delivers only about $40\,\%$ of the rate it would at the surface
concentration throughout; the inner core is wasted catalyst. The cure is
to shrink the pellet (cutting $\phi$ linearly) or to open the pore
structure (raising $D_e$), each of which pushes $\phi$ back toward the
reaction-limited plateau of figure~\ref{fig:vol04ch07:effectiveness}, at
the cost of higher pressure drop through a bed of smaller particles. The
same effectiveness curve governs gas-solid adsorption kinetics and the
sizing of any porous contactor where reaction or uptake competes with
intraparticle diffusion.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-effectiveness.tex}

\paragraph{Number of transfer units for a curved equilibrium line.} The
closed-form $N_{OG}$ of the scrubber case study assumes a straight
equilibrium line; when the equilibrium curve bends, the integral
$N_{OG} = \int \mathrm{d}Y/(Y - Y^{*})$ must be evaluated numerically, and
the honest way to do it on paper is Simpson's rule on a short table. We
strip a volatile organic from water in a packed column from an inlet gas
mole fraction $y_{N+1} = 0.020$ to an outlet $y_1 = 0.002$, with a
liquid-side equilibrium that curves: the gas mole fraction in equilibrium
with the local liquid follows $y^{*} = 0.6\,x + 8\,x^2$ over the dilute
range, and the operating line from the column material balance is
$y = 0.25\,x + 0.0008$. At five gas compositions spanning the column we
read the local liquid from the operating line, compute $y^{*}$, and form
the integrand $1/(y - y^{*})$:
\[
\begin{array}{lccccc}
y & 0.0020 & 0.0065 & 0.0110 & 0.0155 & 0.0200\\
x = (y-0.0008)/0.25 & 0.0048 & 0.0228 & 0.0408 & 0.0588 & 0.0768\\
y^{*} = 0.6x + 8x^2 & 0.0031 & 0.0178 & 0.0378 & 0.0629 & 0.0933\\
1/(y - y^{*}) & {-909} & {-885} & {-746} & {-633} & {-528}
\end{array}
\]
The integrand is negative because the operating line lies below the
equilibrium line here, which signals that this column is stripping the
solute out of the liquid rather than absorbing it; taking the magnitude and
integrating from $y_1$ to $y_{N+1}$ with Simpson's rule on the five points
(panel width $h = 0.0045$),
\[
N_{OG} = \frac{h}{3}\big[f_0 + 4f_1 + 2f_2 + 4f_3 + f_4\big] =
\frac{0.0045}{3}\big[909 + 4(885) + 2(746) + 4(633) + 528\big].
\]
The bracket is $909 + 3540 + 1492 + 2532 + 528 = 8001$, so
$N_{OG} = (0.0015)(8001) = 12.0$. The curved equilibrium line demands
twelve transfer units, far more than the straight-line estimate of about
five that the same end compositions would give if the curvature were
ignored, because the equilibrium line closes on the operating line in the
concentrated end of the column and starves the driving force exactly where
the load is heaviest. Sizing this column on the linear formula would
underbuild it by more than half its required height, the standard penalty
for treating a bent equilibrium line as straight.

\paragraph{Penetration-theory coefficient from a contact time.} A
gas-liquid contactor renews its liquid surface as bubbles rise through it,
so each liquid element faces the gas only for a short contact time before
it is mixed back into the bulk. Penetration theory turns that contact time
into a mass-transfer coefficient. We size $k_L$ for carbon dioxide
absorbing into water in a bubble column where the rise of a bubble exposes
a fresh liquid surface for an estimated contact time $t_c = 0.05\,\text{s}$,
with $D_{\text{CO}_2} = 1.9\times 10^{-9}\,\si{\meter\squared\per\second}$.
The Higbie result derived in section~7.1 gives the time-averaged
coefficient,
\[
k_L = 2\sqrt{\frac{D}{\pi t_c}} = 2\sqrt{\frac{1.9\times 10^{-9}}
{\pi(0.05)}} = 2\sqrt{1.21\times 10^{-8}} = 2(1.10\times 10^{-4}) =
2.2\times 10^{-4}\,\text{m/s}.
\]
A liquid-side coefficient near $2\times 10^{-4}\,\text{m/s}$ is the working
value for a gas bubbling through water, an order of magnitude above what
pure molecular diffusion through a thick stagnant film would give, because
the surface renewal keeps resetting the penetrating front before it can
flatten. The $\sqrt{D}$ scaling is the diagnostic signature of a
penetration-controlled coefficient, against the $D^{1}$ of a steady film;
when a measured $k_L$ scales as the square root of the diffusivity across a
family of solutes, the contactor is renewing its surface, and the contact
time, not a fixed film thickness, sets the rate.

\paragraph{Cross-current against counter-current extraction.} The contact
pattern, not just the stage count, decides how much solvent an extraction
spends to hit a recovery target. We compare a three-stage cross-current
cascade against a three-stage counter-current cascade for the same duty:
recover acetic acid from water into an organic solvent with partition
coefficient $K = y/x = 2.5$, treating $1.0\,\text{mol/s}$ of feed solute
on a dilute basis. In cross-current operation each stage gets a fresh
slug of solvent; with a per-stage solvent-to-feed ratio that makes the
per-stage extraction factor $E = K(S/F) = 1.0$, the fraction of solute
remaining after each stage is $1/(1+E) = 0.5$, so three stages leave
$0.5^3 = 0.125$, a $87.5\,\%$ recovery, and the total solvent spent is
three full slugs. In counter-current operation the same total solvent
contacts the feed in series, fresh solvent meeting the most-depleted
raffinate; with an overall extraction factor $E = 1.0$ the Kremser form
for fresh solvent leaves a raffinate fraction
\[
\frac{E - 1}{E^{N+1} - 1}\bigg|_{E\to 1} = \frac{1}{N+1} = \frac{1}{4} =
0.25
\]
at three stages only if the solvent were split per stage; using the full
solvent counter-currently the effective extraction factor rises, and at
$E = 1$ the counter-current raffinate fraction is $1/(N+1) = 0.25$ for the
same total solvent that cross-current spends to reach $0.125$. The
comparison cuts the other way once the solvent is scarce: at a tight total
solvent budget giving $E = 1$, counter-current and cross-current land near
each other at three stages, but at $E$ comfortably above one counter-current
pulls decisively ahead, because every stage works against the freshest
solvent it can see. The engineering rule is that counter-current spends
solvent most efficiently and is the default for a continuous plant, while
cross-current, simpler to operate as a batch of separate mixer-settlers,
pays a solvent penalty that grows with the recovery demanded.

\paragraph{Uphill diffusion in a ternary mixture.} Fick's single-coefficient
law predicts that every species moves down its own concentration gradient;
in a mixture of three or more components a species can move up its own
gradient, dragged by the others, and the Maxwell-Stefan frame of the
section~7.1 mastery box is needed to see it. We make the effect concrete
for the classic Duncan-Toor experiment geometry: hydrogen (1), nitrogen
(2), and carbon dioxide (3) in a connecting tube, with nitrogen initially
uniform so that its own gradient is zero, $\nabla x_2 = 0$. Fick would
predict no nitrogen flux. The Maxwell-Stefan relation for nitrogen reads
\[
\nabla x_2 = \frac{x_1 x_2}{D_{12}}\,(v_1 - v_2)\,\frac{1}{c} +
\frac{x_2 x_3}{D_{23}}\,(v_3 - v_2)\,\frac{1}{c},
\]
in molar-flux form $0 = (x_2 N_1 - x_1 N_2)/D_{12} + (x_2 N_3 - x_3 N_2)/
D_{23}$ when $\nabla x_2 = 0$. With hydrogen diffusing fast against carbon
dioxide ($D_{12}$ for H$_2$-N$_2$ near $7.8\times 10^{-5}$ and $D_{23}$ for
CO$_2$-N$_2$ near $1.7\times 10^{-5}\,\si{\meter\squared\per\second}$, a
factor of about $4.6$ apart), the two friction terms cannot both vanish, so
$N_2 \neq 0$. Solving the balance for $N_2$ at a representative
$x_1 = x_2 = x_3 = 1/3$,
\[
N_2 = \frac{x_2\,(N_1/D_{12} + N_3/D_{23})}{x_1/D_{12} + x_3/D_{23}},
\]
and with hydrogen and carbon dioxide counter-diffusing ($N_1 = -N_3 = N$
for equimolar exchange) the numerator is $x_2 N(1/D_{12} - 1/D_{23})$, which
is non-zero and negative because $D_{23} < D_{12}$. Nitrogen is therefore
pushed in the direction set by the carbon-dioxide friction, against its own
flat profile, and over the transient it develops a concentration bump where
none should exist by Fick's law. The lesson for design is direct: in
multicomponent distillation, in membrane gas separation with three or more
permeants, and in any ternary diffusion problem, the single-diffusivity
Fick model gives qualitatively wrong fluxes, and the pairwise
Maxwell-Stefan diffusivities must be carried.

\paragraph{Equimolar counterdiffusion against Stefan diffusion, side by
side.} The two binary diffusion modes give different fluxes for the same
end compositions, and putting the numbers next to each other fixes how
large the gap grows with concentration. Ammonia diffuses across a gas film
of thickness $L = 2.0\,\si{\milli\meter}$ at $25\,\degC$ and $1\,\text{atm}$
with $D_{AB} = 2.3\times 10^{-5}\,\si{\meter\squared\per\second}$, from a
face where its mole fraction is $x_{A,1} = 0.40$ to a face where
$x_{A,2} = 0.10$. The total concentration is $c = p/(RT) = 101\,325/
[(8.314)(298)] = 40.9\,\text{mol/m}^3$. In equimolar counterdiffusion the
partner gas moves back through the film at the same rate, the bulk flow
cancels, and the flux is the bare Fick value,
\[
N_A^{\text{equi}} = \frac{c\,D_{AB}(x_{A,1}-x_{A,2})}{L} =
\frac{(40.9)(2.3\times 10^{-5})(0.30)}{2.0\times 10^{-3}} =
1.41\times 10^{-1}\,\text{mol/m}^2\text{s}.
\]
In Stefan diffusion through a stagnant partner the log-mean inert fraction
enters. The inert mole fractions are $x_{B,1} = 0.60$ and $x_{B,2} = 0.90$,
so $x_{B,\text{lm}} = (0.90-0.60)/\ln(0.90/0.60) = 0.30/0.405 = 0.740$, and
\[
N_A^{\text{Stefan}} = \frac{c\,D_{AB}(x_{A,1}-x_{A,2})}{L\,x_{B,\text{lm}}}
= \frac{1.41\times 10^{-1}}{0.740} = 1.91\times 10^{-1}\,\text{mol/m}^2
\text{s}.
\]
Stefan diffusion delivers $35\,\%$ more flux than equimolar
counterdiffusion at these compositions, because the diffusing ammonia
drags the stagnant gas it pushes into and that induced convection adds to
the transport. The gap is the factor $1/x_{B,\text{lm}} = 1.35$. Repeating
the arithmetic for a dilute case, $x_{A,1} = 0.04$ and $x_{A,2} = 0.01$,
gives $x_{B,\text{lm}} = 0.975$ and a gap of only $2.5\,\%$, the
quantitative statement of why the two modes are indistinguishable for a
trace species and must be told apart for a concentrated one. The working
rule: a distillation tray, where heavy condenses as light vapourises, is
equimolar and takes bare Fick; an evaporating or absorbing film against an
insoluble carrier is Stefan and takes the log-mean factor.

\paragraph{Mass-transfer coefficient from heat-transfer data by the
Colburn analogy.} The analogy earns its keep when a geometry has a
published heat-transfer correlation and no mass-transfer one. Air at
$25\,\degC$ flows over a naphthalene casting at conditions where a
heat-transfer test on the identical geometry reports a heat-transfer
coefficient $h = 45\,\text{W}/(\text{m}^2\cdot\text{K})$. We want the
mass-transfer coefficient $k_c$ for naphthalene subliming into the air,
with air properties $\rho = 1.18\,\text{kg/m}^3$, $c_p = 1005\,\text{J}/
(\text{kg}\cdot\text{K})$, $\mathrm{Pr} = 0.71$, and a naphthalene-in-air
diffusivity $D_{AB} = 0.62\times 10^{-5}\,\si{\meter\squared\per\second}$
giving $\mathrm{Sc} = \nu/D_{AB} = (1.55\times 10^{-5})/(0.62\times
10^{-5}) = 2.50$. Equate the Colburn factors $j_H = j_D$,
\[
\frac{h}{\rho c_p u}\,\mathrm{Pr}^{2/3} = \frac{k_c}{u}\,\mathrm{Sc}^{2/3}
\;\Rightarrow\; k_c = \frac{h}{\rho c_p}
\left(\frac{\mathrm{Pr}}{\mathrm{Sc}}\right)^{2/3}.
\]
The velocity cancels, which is the convenience of the $j$-factor form. The
prefactor is $h/(\rho c_p) = 45/[(1.18)(1005)] = 3.79\times 10^{-2}\,
\text{m/s}$, and the diffusivity-ratio factor is $(\mathrm{Pr}/\mathrm{Sc})
^{2/3} = (0.71/2.50)^{2/3} = (0.284)^{2/3} = 0.434$, so
\[
k_c = (3.79\times 10^{-2})(0.434) = 1.65\times 10^{-2}\,\text{m/s}.
\]
A coefficient near $1.6\,\text{cm/s}$ for the subliming surface, obtained
without a single mass-transfer measurement. The factor
$(\mathrm{Pr}/\mathrm{Sc})^{2/3} < 1$ here because the Schmidt number
exceeds the Prandtl number: the species boundary layer is thinner than the
thermal one, so per unit of analogous driving force the mass transfer is
slower. The estimate is good wherever the heat-transfer correlation came
from the same flow regime; a heat correlation contaminated by radiation or
by a wall-temperature-dependent property would carry that error into
$k_c$.

\paragraph{Surface-renewal rate from a measured liquid-side coefficient.}
The three interfacial models predict different scalings, and a measured
coefficient can be read back into a renewal rate to check which model the
contactor obeys. A stirred gas-liquid cell absorbing carbon dioxide into
water at $25\,\degC$ reports a liquid-side coefficient $k_L = 3.0\times
10^{-4}\,\text{m/s}$, with $D_{\text{CO}_2} = 1.9\times 10^{-9}\,\si{
\meter\squared\per\second}$. Reading it through the Danckwerts
surface-renewal result $k_L = \sqrt{D\,s}$ gives the renewal rate
\[
s = \frac{k_L^2}{D} = \frac{(3.0\times 10^{-4})^2}{1.9\times 10^{-9}} =
\frac{9.0\times 10^{-8}}{1.9\times 10^{-9}} = 47\,\text{s}^{-1},
\]
so each patch of surface survives on average $1/s = 0.021\,\text{s}$
before bulk fluid replaces it. Reading the same coefficient through the
Higbie penetration result $k_L = 2\sqrt{D/(\pi t_c)}$ instead gives a
single contact time $t_c = 4D/(\pi k_L^2) = 4(1.9\times 10^{-9})/
[\pi(9.0\times 10^{-8})] = 0.027\,\text{s}$, the same order as the mean
renewal age, as it must be since the two models share the square-root
physics. Reading it through the film model $k_L = D/\delta$ would give a
film thickness $\delta = D/k_L = 1.9\times 10^{-9}/3.0\times 10^{-4} =
6.3\,\si{\micro\meter}$, a plausible number but an unmeasurable one. The
discriminating test is to repeat the measurement for a second solute,
oxygen at $D = 2.1\times 10^{-9}$: surface renewal predicts $k_L$ rises as
$\sqrt{D}$, by the factor $\sqrt{2.1/1.9} = 1.05$, while the film model
predicts it rises as $D^{1}$, by $2.1/1.9 = 1.11$. A coefficient that
moves by five percent rather than eleven confirms the renewing surface and
fixes $s$ as the controlling parameter.

\paragraph{Tissue oxygen reach by the Krogh-cylinder balance.} Oxygen
diffusing out of a capillary into the tissue it feeds reaches only a
finite radius before metabolism consumes it, and the Krogh-cylinder model
sets that radius from a steady diffusion-with-consumption balance. Each
capillary supplies a coaxial cylinder of tissue of radius $R$, inside
which oxygen diffuses radially and is consumed at a constant volumetric
rate $M_0$ (zero-order metabolism). The steady species balance in
cylindrical coordinates is
\[
\frac{D_t}{r}\frac{\mathrm{d}}{\mathrm{d}r}\!\left(r\frac{\mathrm{d}C}
{\mathrm{d}r}\right) = M_0,
\]
with $D_t$ the oxygen diffusivity in tissue. Integrating twice with a
fixed capillary-wall concentration $C_w$ at $r = r_c$ and a no-flux outer
boundary $\mathrm{d}C/\mathrm{d}r = 0$ at $r = R$ (the edge of the
cylinder, where the next capillary's territory begins) gives the lethal
corner condition that the concentration just reaches zero at $r = R$ when
the cylinder is sized to the supply. Solving for the reach with $C(R) = 0$,
\[
C_w = \frac{M_0}{4 D_t}\!\left[R^2\ln\!\frac{R}{r_c} - \tfrac{1}{2}
(R^2 - r_c^2)\right]\Big/\,\text{(leading term)} \approx
\frac{M_0 R^2}{4 D_t}\ln\!\frac{R}{r_c},
\]
the working form when $R \gg r_c$. Putting in muscle numbers, $C_w =
4.0\times 10^{-2}\,\text{mol/m}^3$ (the dissolved oxygen at the wall),
$M_0 = 1.5\times 10^{-2}\,\text{mol}/(\text{m}^3\cdot\text{s})$,
$D_t = 1.7\times 10^{-9}\,\si{\meter\squared\per\second}$, and $r_c =
4\,\si{\micro\meter}$, and solving $M_0 R^2 \ln(R/r_c) = 4 D_t C_w$
iteratively: trying $R = 30\,\si{\micro\meter}$, the left side is
$(1.5\times 10^{-2})(9.0\times 10^{-10})\ln(7.5) = (1.35\times 10^{-11})
(2.01) = 2.7\times 10^{-11}$ against a right side $4(1.7\times 10^{-9})
(4.0\times 10^{-2}) = 2.7\times 10^{-10}$, low by ten, so $R$ must rise;
trying $R = 90\,\si{\micro\meter}$ gives left side $(1.5\times 10^{-2})
(8.1\times 10^{-9})\ln(22.5) = (1.22\times 10^{-10})(3.11) = 3.8\times
10^{-10}$, now above, so the reach sits near $R \approx 80\,\si{
\micro\meter}$. Capillaries spaced about $160\,\si{\micro\meter}$ apart
just keep muscle tissue oxygenated, which matches the measured capillary
density of working muscle and explains why a tissue whose metabolism rises
or whose capillaries rarefy develops a hypoxic core. The same balance
sizes the fibre spacing in the blood oxygenator of the case study below,
with the diffusion running the other way.

\paragraph{Dialysis clearance from an overall membrane coefficient.} A
membrane separating two flowing liquids transfers solute by the same
two-resistance series as a gas-liquid film, and a haemodialyser is the
clean example. Urea crosses a dialysis membrane from blood into dialysate,
with a blood-side film coefficient $k_b = 4.0\times 10^{-5}\,\text{m/s}$,
a membrane permeance $k_m = 2.5\times 10^{-5}\,\text{m/s}$, and a
dialysate-side coefficient $k_d = 6.0\times 10^{-5}\,\text{m/s}$, the three
in series. The overall coefficient is
\[
\frac{1}{K_o} = \frac{1}{k_b} + \frac{1}{k_m} + \frac{1}{k_d} =
\frac{1}{4.0\times 10^{-5}} + \frac{1}{2.5\times 10^{-5}} +
\frac{1}{6.0\times 10^{-5}},
\]
the resistances $2.50\times 10^{4} + 4.00\times 10^{4} + 1.67\times 10^{4}
= 8.17\times 10^{4}\,\text{s/m}$, so $K_o = 1.22\times 10^{-5}\,\text{m/s}$.
The membrane itself carries $4.00/8.17 = 49\,\%$ of the resistance and
sets the clearance; thinning the membrane or opening its pores is the
direct lever, while improving either flow film changes the rate little.
For a dialyser of membrane area $A = 1.8\,\text{m}^2$, the mass-transfer
capacity is $K_o A = (1.22\times 10^{-5})(1.8) = 2.2\times 10^{-5}\,
\text{m}^3/\text{s} = 1.3\,\text{L/min}$, which against a blood flow of
$0.3\,\text{L/min}$ gives a clearance close to the blood flow itself, the
flow-limited regime a clinical dialyser is designed to reach. The number
that matters to the patient, the urea clearance, falls straight out of the
slowest film in the series, the same principle the chapter's closing
principle states for every multi-phase contactor.

\paragraph{Developing mass transfer in a tube: the entry-length coefficient.}
A short wetted-wall or hollow-fibre contactor works in the entry region,
where the concentration boundary layer has not yet filled the tube and the
local coefficient runs several times its fully developed value; sizing such a
device on the fully developed asymptote underpredicts its transfer. We put a
number on the gap for laminar flow of a liquid through a fibre of inside
diameter $d = 0.4\,\si{\milli\meter}$ and length $\ell = 0.20\,\si{\meter}$,
carrying a solute of diffusivity $D_{AB} = 1.0\times 10^{-9}\,\si{
\meter\squared\per\second}$ at a mean velocity $u = 0.05\,\text{m/s}$, with
kinematic viscosity $\nu = 1.0\times 10^{-6}\,\si{\meter\squared\per\second}$.
The Reynolds number is $\text{Re} = ud/\nu = (0.05)(4.0\times 10^{-4})/
(1.0\times 10^{-6}) = 20$, deep in the laminar range, and the Schmidt number
is $\text{Sc} = \nu/D_{AB} = 1.0\times 10^{-6}/1.0\times 10^{-9} = 1000$, the
large value that liquids carry. The dimensionless length that decides whether
the flow has developed is the inverse Graetz number,
\[
x^{*} = \frac{\ell/d}{\text{Re}\,\text{Sc}} = \frac{0.20/(4.0\times 10^{-4})}
{(20)(1000)} = \frac{500}{2.0\times 10^{4}} = 2.5\times 10^{-2}.
\]
A value well below the $0.05$ that marks the end of the concentration entry
length means the fibre lives inside the developing region, so the mean
Sherwood number is set by the Leveque entry solution rather than the
constant-wall-concentration asymptote of $3.66$. The Leveque mean-Sherwood
result for a tube with a developing concentration layer is $\text{Sh} =
1.62\,(x^{*})^{-1/3}$, so
\[
\text{Sh} = 1.62\,(2.5\times 10^{-2})^{-1/3} = 1.62\,(3.42) = 5.5,
\]
half again above the fully developed $3.66$. The mass-transfer coefficient is
$k_c = \text{Sh}\,D_{AB}/d = (5.5)(1.0\times 10^{-9})/(4.0\times 10^{-4}) =
1.4\times 10^{-5}\,\text{m/s}$, against the $9.2\times 10^{-6}\,\text{m/s}$
the asymptote would give. Figure~\ref{fig:vol04ch07:entry-length} draws the
local Sherwood number decaying down the tube toward $3.66$ and shows the
inlet region carrying the larger coefficient. The design reading is that a
mass-transfer device deliberately kept short, so that the flow never fully
develops, runs at a coefficient the fully developed formula would miss, which
is one reason hollow-fibre and wetted-wall geometries are built with many
short parallel passages rather than a few long ones.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-entry-length.tex}

\paragraph{Diffusion time through a polymer film: permeation lag.} A gas
diffusing through a dense polymer membrane does not appear on the downstream
side the instant the upstream pressure is raised; it appears after a transient
during which the film loads up its dissolved-gas profile, and the transient
has a clean signature, the time lag, that reads the diffusivity straight off a
permeation experiment. For a film of thickness $L$ held at a fixed upstream
concentration from $t = 0$ with a downstream side kept near zero, the
downstream cumulative flux approaches a straight line whose backward extension
crosses the time axis at the time lag
\[
t_{\text{lag}} = \frac{L^2}{6\,D},
\]
the standard result of the transient diffusion equation with these boundary
conditions. We size the lag for oxygen through a food-packaging film of
thickness $L = 50\,\si{\micro\meter}$ with an oxygen diffusivity in the
polymer $D = 2.0\times 10^{-12}\,\si{\meter\squared\per\second}$,
\[
t_{\text{lag}} = \frac{(5.0\times 10^{-5})^2}{6\,(2.0\times 10^{-12})} =
\frac{2.5\times 10^{-9}}{1.2\times 10^{-11}} = 208\,\text{s} \approx
3.5\,\text{min}.
\]
The steady permeation then follows at the permeability times the
pressure difference. The time lag is the working way to separate the two
factors that make up the permeability $P = D\,S$: the steady flux fixes the
product $D\,S$, and the lag fixes $D$ alone, so the solubility $S$ falls out
by division. For the packaging engineer the lag is usually a curiosity next
to the steady rate, three minutes against a shelf life of months, but for a
thick barrier or a slow-diffusing penetrant it can stretch to hours and
governs the response time of a membrane sensor, where the device cannot read
a step change in feed until the film has loaded through this same $L^2/6D$
transient.

\paragraph{Adsorption bed depth from the mass-transfer zone.} A fixed-bed
adsorber is sized so the mass-transfer zone, the travelling band of
partly loaded solid where the concentration front lives, fits inside the bed
with the fully loaded section ahead of it doing the work. We size an
activated-carbon bed that strips a volatile organic from an air stream, given
a measured breakthrough run on a short test column. The test column of length
$L_{\text{test}} = 0.30\,\text{m}$ shows the outlet concentration first
break at $t_b = 4.0\,\text{h}$ and reach the feed value (exhaustion) at
$t_e = 5.2\,\text{h}$ at the same superficial velocity the full bed will run.
The fraction of the bed that is the still-unused mass-transfer zone at
breakthrough is set by how much capacity remains between break and exhaustion,
\[
f_{\text{unused}} = \frac{1}{2}\,\frac{t_e - t_b}{t_e} =
\frac{1}{2}\,\frac{5.2 - 4.0}{5.2} = \frac{1}{2}(0.231) = 0.115,
\]
taking the front as roughly symmetric so half the zone is unused at
breakthrough. The length of unused bed is then $L_{\text{UB}} =
f_{\text{unused}}\,L_{\text{test}} = 0.115(0.30) = 0.035\,\text{m}$, a fixed
length that travels with the zone and does not change when the bed is made
longer. To run a full bed to a service time of $t_{\text{service}} =
24\,\text{h}$ before switchover, the fully loaded length scales with the
service time at the front velocity that the test fixes, $L_{\text{loaded}} =
L_{\text{test}}(t_{\text{service}}/t_b) - L_{\text{UB}}$ scaled through the
equilibrium section; carrying the zone length as a fixed addition, the
required bed depth is
\[
L_{\text{bed}} = \frac{t_{\text{service}}}{t_b}\,(L_{\text{test}} -
L_{\text{UB}}) + L_{\text{UB}} = \frac{24}{4.0}\,(0.30 - 0.035) + 0.035 =
6(0.265) + 0.035 = 1.62\,\text{m}.
\]
A bed near $1.6\,\text{m}$ deep gives a full day between regenerations, and
the fixed $3.5\,\text{cm}$ of unused bed is the small penalty a sharp
favourable-isotherm front costs. Had the isotherm been unfavourable, the zone
would spread as it travelled, $L_{\text{UB}}$ would grow with bed length
rather than staying fixed, and the bed would waste a rising fraction of its
capacity, the case the isotherm-shape discussion of section~7.5 warns against.

\paragraph{Simultaneous heat and mass transfer: the wet-bulb temperature.}
A wet surface evaporating into unsaturated air cools itself, because the
latent heat of the departing vapour is drawn from the surface, and it settles
at the wet-bulb temperature where the sensible heat arriving from the air
exactly balances the latent heat leaving with the vapour. This steady balance
is the constant-rate drying temperature and the reading of a psychrometer, and
it couples the heat-transfer coefficient $h$ to the mass-transfer coefficient
$k_c$ through the Lewis number. The balance sets convective heat gain equal to
evaporative heat loss,
\[
h\,(T_\infty - T_w) = k_c\,M_A\,h_{fg}\,(c_{A,w} - c_{A,\infty}),
\]
with $T_w$ the wet-bulb temperature, $c_{A,w}$ the saturation vapour
concentration at $T_w$, and $M_A h_{fg}$ the latent heat per mole. We solve
for the wet-bulb depression of air at $T_\infty = 40\,\degC$ and $20\,\%$
relative humidity. The Chilton-Colburn analogy fixes the coefficient ratio,
$h/k_c = \rho c_p\,\text{Le}^{2/3}$, and for the air-water system the Lewis
number is near unity, $\text{Le} = \alpha/D_{AB} \approx 0.87$, so
$\text{Le}^{2/3} = 0.91$ and $h/k_c = (1.14)(1005)(0.91) = 1043\,\text{J}/
(\text{m}^3\cdot\text{K})$. The saturation vapour concentration at $40\,\degC$
is $c_{\text{sat}}(40) = p_{\text{sat}}/(RT) = 7384/[(8.314)(313)] =
2.84\,\text{mol/m}^3$; the air carries $0.20$ of the value it would hold
saturated, $c_{A,\infty} = 0.20(2.84) = 0.57\,\text{mol/m}^3$. Guessing a
wet-bulb near $23\,\degC$, the saturation concentration there is
$c_{\text{sat}}(23) = 2810/[(8.314)(296)] = 1.14\,\text{mol/m}^3$. The mass
driving force is $c_{A,w} - c_{A,\infty} = 1.14 - 0.57 = 0.57\,\text{mol/m}^3$,
and with $M_A h_{fg} = (0.018)(2.44\times 10^{6}) = 4.39\times 10^{4}\,
\text{J/mol}$ the required temperature depression is
\[
T_\infty - T_w = \frac{k_c\,M_A h_{fg}\,(c_{A,w}-c_{A,\infty})}{h} =
\frac{M_A h_{fg}\,(c_{A,w}-c_{A,\infty})}{\rho c_p\,\text{Le}^{2/3}} =
\frac{(4.39\times 10^{4})(0.57)}{1043} = 24\,\degC.
\]
That gives $T_w = 40 - 24 = 16\,\degC$, below the $23\,\degC$ guess, so a
second pass with a lower trial wet-bulb (which lowers $c_{A,w}$ and shrinks
the depression) converges near $T_w \approx 21\,\degC$, a $19\,\degC$
depression, close to the psychrometric-chart reading for these conditions.
The lesson the arithmetic carries is that the wet-bulb temperature is a
mass-transfer quantity in heat-transfer clothing: it exists because the same
boundary layer carries both fluxes, and the near-unity Lewis number of the
air-water system is exactly why the wet-bulb temperature and the
adiabatic-saturation temperature nearly coincide for that one pair, a
coincidence that fails for other solvent-gas systems where $\text{Le}$
departs from one.

\paragraph{Osmotic pressure and the minimum work of separation.} Every
separation has a thermodynamic floor, the minimum reversible work to pull the
product out of the feed, and no real device can beat it; comparing the actual
energy to the floor is the honest efficiency of a separation. We compute the
floor for desalting seawater and set it against the working figure of the
reverse-osmosis case study. The minimum work to produce a small permeate
fraction from a large feed, in the limit of vanishing recovery so the feed is
undepleted, is the osmotic pressure of the feed times the permeate volume,
\[
w_{\min} = \pi_f\,V_p \quad\Rightarrow\quad
\frac{w_{\min}}{V_p} = \pi_f = 2.7\,\text{MPa} = 2.7\times 10^{6}\,
\text{J/m}^3.
\]
Converting to the working unit, $2.7\times 10^{6}\,\text{J/m}^3 =
(2.7\times 10^{6})/(3.6\times 10^{6})\,\text{kWh/m}^3 = 0.75\,\text{kWh/m}^3$.
At a finite recovery the floor rises, because the brine left behind is
saltier and holds its water more tightly; for the case study's $r = 0.45$ the
reversible minimum, integrated over the concentrating brine, is near
$1.1\,\text{kWh/m}^3$. The reverse-osmosis skid of the second case study runs
at about $3\,\text{kWh/m}^3$ with energy recovery, so its second-law
efficiency is roughly $1.1/3 \approx 0.37$, and the pre-energy-recovery
figure of $6\,\text{kWh/m}^3$ sat near $0.18$. The gap from the floor is
spent on the finite driving pressure a real membrane needs to pass water at a
useful flux, on the pump and pressure-exchanger inefficiencies, and on
concentration polarisation, in that order. The number worth carrying is that
seawater desalination has closed to within a factor of about three of its
thermodynamic minimum, which is why the field has stopped chasing large
energy gains from the separation itself and now chases them in pretreatment,
membrane fouling, and the recovery device, where the remaining slack lives.

\paragraph{Gas diffusivity from Chapman-Enskog, and its correction to a new
state.} A diffusivity read from the room-temperature table at the head of the
chapter is rarely at the exact temperature and pressure a process runs, and
the Chapman-Enskog scaling of the section's mastery box corrects it in two
steps. We carry oxygen in nitrogen from its tabulated $D = 2.0\times 10^{-5}
\,\si{\meter\squared\per\second}$ at $298\,\si{\kelvin}$ and $1\,\text{atm}$
to a furnace preheater running at $500\,\si{\kelvin}$ and $2\,\text{atm}$. The
kinetic-theory result gives $D_{AB} \propto T^{3/2}/p$, so the corrected
diffusivity is the tabulated value times the temperature ratio to the
three-halves power and the inverse pressure ratio,
\[
D_{AB}(500,2) = (2.0\times 10^{-5})\left(\frac{500}{298}\right)^{3/2}
\left(\frac{1}{2}\right).
\]
The temperature factor is $(1.678)^{3/2} = \exp(1.5\ln 1.678) =
\exp(1.5\cdot 0.5175) = \exp(0.776) = 2.17$, and the pressure factor is
$0.5$, so $D_{AB} = (2.0\times 10^{-5})(2.17)(0.5) = 2.2\times 10^{-5}\,\si{
\meter\squared\per\second}$, close to the starting value because the two
corrections nearly cancel. The near-cancellation is worth carrying as a
warning: a preheated, pressurised gas stream can have a diffusivity within
ten percent of its room-temperature value even though both the temperature
and the pressure have moved substantially, because the two effects oppose.
The collision integral $\Omega_D$ that the full formula carries drifts only
weakly with temperature and shifts the answer by a few percent at most over
this range, which is why the bare $T^{3/2}/p$ scaling is the working correction
and the collision integral is the refinement.
Figure~\ref{fig:vol04ch07:diffusivity-temperature} draws the scaling and
marks how a factor of several separates the low- and high-temperature ends of
the plotted range \cite{text:bird-transport-phenomena}.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-diffusivity-temperature.tex}

\paragraph{Kremser stage count inverted at a target recovery.} The Kremser
relation is most useful run backward: given a required recovery and an
absorption factor, solve for the stage count a column needs. We size the
stages for an absorber that must recover $98\,\%$ of a solute from a gas into
a lean solvent, with an absorption factor $A = 1.5$ set by the chosen
solvent-to-gas ratio and the equilibrium slope. The Kremser recovery with
lean entering solvent is
\[
\phi = \frac{A^{N+1} - A}{A^{N+1} - 1} = 0.98.
\]
Solving for $N$, rearrange to $A^{N+1}(1 - \phi) = A - \phi$, so $A^{N+1} =
(A - \phi)/(1 - \phi) = (1.5 - 0.98)/(1 - 0.98) = 0.52/0.02 = 26.0$, and
\[
N + 1 = \frac{\ln 26.0}{\ln 1.5} = \frac{3.258}{0.405} = 8.04,
\]
so $N = 7.0$ equilibrium stages. Tightening the target to $99\,\%$ raises the
right side to $(1.5 - 0.99)/0.01 = 51.0$ and $N + 1 = \ln 51/\ln 1.5 =
9.66$, so $N = 8.7$, nearly two more stages for the last percent of recovery,
the steep-cost signature of chasing a high recovery that
figure~\ref{fig:vol04ch07:kremser-recovery} draws as the flattening approach
to the recovery ceiling. Raising the absorption factor is the cheaper lever:
at $A = 2.0$ the $98\,\%$ target needs $A^{N+1} = (2.0 - 0.98)/0.02 = 51.0$
and $N + 1 = \ln 51/\ln 2 = 5.67$, so $N = 4.7$ stages, a third fewer than at
$A = 1.5$. The design reading is that more solvent (a higher $A$) and more
stages (a higher $N$) both buy recovery, and the economic choice trades the
solvent circulation and its regeneration duty against the column height, the
same stages-against-reflux trade that distillation runs
\cite{text:treybal-mass-transfer}.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-kremser-recovery.tex}

\paragraph{The economic reflux from the stages-against-reflux trade.} A
distillation column can be built tall with little reflux or short with much,
and the total cost has a minimum between the two limits. We locate the
economic reflux for a separation whose minimum reflux is $R_{\min} = 1.44$
and whose total-reflux minimum stage count is $N_{\min} = 6.4$, the numbers of
the chapter's benzene-toluene and estimation examples. The stage count near
the minimum reflux follows a hyperbola in the reflux, well modelled as
$N = N_{\min} + K/(R - R_{\min})$ with $K$ a fitting constant; taking $K =
3.2$ from a single Gilliland point (at $R = 1.88$, $N \approx 13$), the stage
count at a candidate reflux is set. The capital cost tracks the stage count
(shell and internals) while the operating cost tracks the reflux (reboiler
and condenser duty, both scaling with the vapour rate $V = (R+1)D$). Writing
a dimensionless total cost as the sum of a stage term and a reflux term,
$C = N + \beta(R+1)$ with $\beta$ the relative price of a unit of duty
against a stage, and minimising over $R$,
\[
\frac{\mathrm{d}C}{\mathrm{d}R} = -\frac{K}{(R - R_{\min})^2} + \beta = 0
\;\Rightarrow\; R_{\text{opt}} = R_{\min} + \sqrt{K/\beta}.
\]
For a duty-to-stage price ratio $\beta = 4$, $R_{\text{opt}} = 1.44 +
\sqrt{3.2/4} = 1.44 + 0.894 = 2.33$, about $1.6\times R_{\min}$; for a
cheaper-energy plant with $\beta = 1$, $R_{\text{opt}} = 1.44 + 1.79 = 3.23$,
about $2.2\times R_{\min}$. The optimum shifts toward more reflux, and fewer
stages, when energy is cheap and toward more stages when energy is dear,
which is the quantitative statement behind the working rule of thumb that
columns run at $1.2$ to $1.5$ times the minimum reflux in an energy-priced
world. Figure~\ref{fig:vol04ch07:reflux-stages} draws the hyperbola with its
two asymptotes and marks the economic band at the knee of the curve. The
lesson is that the reflux ratio is not a free choice but the outcome of a cost
minimisation, and that the fashionable band of $1.2$ to $1.5$ times minimum is
the band where energy is neither free nor ruinous.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-reflux-stages.tex}

\paragraph{Concentration polarisation: the wall concentration and the flux it
suppresses.} A membrane sized on the bulk osmotic pressure overpredicts its
flux, because the rejected solute piles up at the wall and lifts the local
osmotic pressure above the bulk. We put a number on the effect for a seawater
reverse-osmosis element running a water flux $J_w = 3.1\times 10^{-6}\,
\text{m/s}$ (the log-mean flux of the case study) against a feed-side liquid
mass-transfer coefficient $k = 2.0\times 10^{-5}\,\text{m/s}$ set by the
feed-spacer hydraulics. The polarisation factor is the exponential of the
flux-to-coefficient ratio,
\[
\frac{c_w}{c_b} = \exp\!\left(\frac{J_w}{k}\right) = \exp\!\left(\frac{3.1
\times 10^{-6}}{2.0\times 10^{-5}}\right) = \exp(0.155) = 1.17,
\]
so the wall salt sits $17\,\%$ above the bulk. The wall osmotic pressure rises
in proportion for a dilute solution, from the bulk $2.7\,\text{MPa}$ to
$1.17(2.7) = 3.2\,\text{MPa}$, which trims the net driving pressure at the
feed end from $7.0 - 2.7 = 4.3\,\text{MPa}$ to $7.0 - 3.2 = 3.8\,\text{MPa}$,
an $12\,\%$ cut, and the flux with it. The suppression is self-reinforcing:
the lower flux the polarisation causes would in turn lower the polarisation,
so the operating point settles at a fixed point of the coupling somewhat below
the bare-flux estimate. Halving the mass-transfer coefficient to $1.0\times
10^{-5}\,\text{m/s}$ (a poorly swept channel) doubles the exponent to $0.31$
and raises the wall factor to $\exp(0.31) = 1.36$, a $36\,\%$ wall
enrichment that would eat a quarter of the driving pressure.
Figure~\ref{fig:vol04ch07:polarisation-profile} draws the exponential film for
a well-swept and a poorly swept channel and shows the wall pile-up the
cross-flow must sweep away. The cure is to raise $k$ by faster cross-flow, a
narrower feed channel, or a turbulence-promoting spacer, all of which thin the
film and pull the exponent toward zero; raising the applied pressure only
raises the flux, which raises the polarisation, and buys less than the pump
work it costs.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-polarisation-profile.tex}

\paragraph{Cooling-crystallisation yield and the seed charge.} A batch
crystalliser cooled from a hot charge drops its product as the solubility
falls, and the yield and the seed loading follow from a solute balance across
the cool. We crystallise a salt from water, cooling $1000\,\text{kg}$ of
solution from $60\,\degC$, where the solubility is $c^{*}_{\text{hot}} =
0.55\,\text{kg solute per kg water}$, to $20\,\degC$, where it is
$c^{*}_{\text{cold}} = 0.30$. The charge is saturated hot, holding
$700\,\text{kg}$ of water and $0.55(700) = 385\,\text{kg}$ of dissolved
solute. Cooling to $20\,\degC$ leaves in solution only $0.30(700) =
210\,\text{kg}$ (taking the water inventory as fixed, no evaporation), so the
crystal yield is
\[
Y = (c^{*}_{\text{hot}} - c^{*}_{\text{cold}})\,m_{\text{water}} =
(0.55 - 0.30)(700) = 175\,\text{kg}
\]
of crystals, a $45\,\%$ recovery of the dissolved solute. To hold the
supersaturation inside the metastable band the whole way down, the charge is
seeded, and the seed mass sets the final crystal size. If the product is to
grow only on the seeds with no new nucleation, every seed crystal must grow by
the same volume ratio, so the final-to-seed mass ratio equals the cube of the
final-to-seed size ratio. For a seed charge of small crystals of mean size
$L_s = 100\,\si{\micro\meter}$ grown to a product size $L_p = 500\,\si{
\micro\meter}$, the mass ratio is $(L_p/L_s)^3 = 5^3 = 125$, so the product
mass is $125$ times the seed mass, and the seed charge needed to carry the
$175\,\text{kg}$ of yield onto seeds alone is $m_{\text{seed}} =
(175 + m_{\text{seed}})/125$, giving $m_{\text{seed}} = 175/124 =
1.4\,\text{kg}$ of seed. Charging less than this forces the excess
supersaturation into new nuclei and shifts the product finer; charging more
grows a coarser, cleaner product that filters and washes well.
Figure~\ref{fig:vol04ch07:metastable-zone} draws the cooling path inside the
metastable band against a crash-cool that crosses the nucleation ceiling. The
arithmetic is the reason a crystalliser is seeded and cooled on a slow
programme rather than crashed: the seed charge and the cooling rate together
fix the crystal-size distribution, and the yield is set by the solubility
curve regardless of how the cooling is run \cite{text:perry-chemical-engineers}.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-metastable-zone.tex}

\paragraph{Knudsen diffusion in a fine pore.} When a gas diffuses through a
pore narrower than its mean free path, the molecules strike the pore wall more
often than each other, and the transport switches from ordinary molecular
diffusion to Knudsen diffusion, with a diffusivity set by the pore size rather
than by the gas pair. We compare the two regimes for nitrogen in a catalyst
pore. The Knudsen diffusivity is $D_K = (d_p/3)\sqrt{8RT/(\pi M)}$, the pore
diameter times a third of the mean molecular speed. For a pore of diameter
$d_p = 10\,\si{\nano\meter}$ carrying nitrogen ($M = 0.028\,\text{kg/mol}$) at
$400\,\si{\kelvin}$, the mean speed group is
\[
\sqrt{\frac{8RT}{\pi M}} = \sqrt{\frac{8(8.314)(400)}{\pi(0.028)}} =
\sqrt{\frac{2.66\times 10^{4}}{0.0880}} = \sqrt{3.02\times 10^{5}} =
550\,\text{m/s},
\]
so $D_K = (10\times 10^{-9}/3)(550) = 1.8\times 10^{-6}\,\si{\meter\squared
\per\second}$, more than an order of magnitude below the ordinary gas
diffusivity of $\sim 3\times 10^{-5}\,\si{\meter\squared\per\second}$. The
pore is Knudsen-controlled, and its transport falls as the pore narrows and
rises only as the square root of temperature, not the $T^{3/2}$ of molecular
diffusion, because the wall collisions that now set the rate do not depend on
the gas density. The combined diffusivity across the transition regime adds
the two resistances in the Bosanquet form, $1/D_{\text{eff}} = 1/D_{AB} +
1/D_K$, so the smaller of the two governs; here $D_K$ dominates and
$D_{\text{eff}} \approx D_K$. The lesson for a catalyst designer is that the
effective diffusivity of a fine-pored pellet is set by the pore geometry, not
by the gas pair, and that widening the pores to escape the Knudsen regime
trades intrapellet diffusion against surface area, the same trade the
effectiveness-factor example weighed \cite{text:bird-transport-phenomena}.

\paragraph{Number of transfer units for a stripping column.} A stripper runs
the absorber in reverse, using a gas to pull a volatile solute out of a
liquid, and the transfer-unit count is set by the stripping factor rather than
the absorption factor. We size the transfer units to strip $95\,\%$ of a
dissolved volatile organic from water with air in a packed column, with a
stripping factor $S = mG/L = 2.0$ set by the air-to-water ratio and the
Henry's-law slope. The number of overall liquid-phase transfer units for a
dilute stripper with a linear equilibrium line is the mirror of the absorber
form,
\[
N_{OL} = \frac{S}{S - 1}\,\ln\!\left[\left(1 - \frac{1}{S}\right)
\frac{x_{\text{in}}}{x_{\text{out}}} + \frac{1}{S}\right],
\]
with the liquid stripped from $x_{\text{in}}$ to $x_{\text{out}} =
0.05\,x_{\text{in}}$ for $95\,\%$ removal, so $x_{\text{in}}/x_{\text{out}} =
20$. With $S = 2.0$, $1/S = 0.5$, the bracket is $(1 - 0.5)(20) + 0.5 =
10.5$, and
\[
N_{OL} = \frac{2.0}{1.0}\,\ln(10.5) = 2.0(2.351) = 4.7,
\]
so about five transfer units. Raising the air rate to $S = 3.0$ drops the
bracket to $(1 - 0.333)(20) + 0.333 = 13.7$ and $N_{OL} = (3/2)\ln 13.7 =
1.5(2.617) = 3.9$, fewer units for more air, the stripping analogue of trading
solvent for stages in absorption. A stripping factor at or below unity, like
an absorption factor at or below unity, caps the achievable removal no matter
how tall the column, so an air-stripper is always run with $S$ comfortably
above one, and the air rate is the design lever that the equilibrium slope and
the target removal set together \cite{text:treybal-mass-transfer}.

\paragraph{Single-droplet drying time in a spray dryer.} A spray dryer
atomises the feed into fine droplets whose large surface-to-volume ratio dries
them in seconds, and the residence time the chamber must provide follows from
a single-droplet constant-rate balance. We size the drying time for a milk
droplet of initial diameter $d_0 = 100\,\si{\micro\meter}$ falling through hot
air at a droplet-to-air temperature depression that holds the droplet at its
wet-bulb temperature. In the constant-rate period the droplet evaporates at a
rate set by the gas-side mass transfer over its surface, and the classic
result is the $d^2$-law: the square of the droplet diameter falls linearly
with time, $d^2 = d_0^2 - K t$, with the evaporation constant $K = 8\rho_g
D_{AB}\ln(1 + B_M)/\rho_L$, where $B_M$ is the Spalding mass-transfer number.
For water evaporating into hot dry air, taking $\rho_g D_{AB} \approx
2.5\times 10^{-5}\,\text{kg}/(\text{m}\cdot\text{s})$, $\rho_L = 1000\,\text{
kg/m}^3$, and a mass-transfer number $B_M \approx 0.3$ (moderately dry air),
\[
K = \frac{8(2.5\times 10^{-5})\ln(1.3)}{1000} = \frac{8(2.5\times 10^{-5})
(0.262)}{1000} = 5.2\times 10^{-8}\,\si{\meter\squared\per\second}.
\]
The time to evaporate the free surface water, taking the droplet from $d_0$ to
a dry-shell diameter roughly $0.7\,d_0$ where the falling-rate period begins,
is $t = (d_0^2 - (0.7 d_0)^2)/K = d_0^2(1 - 0.49)/K = (1.0\times 10^{-4})^2
(0.51)/(5.2\times 10^{-8}) = 0.098\,\text{s}$, about a tenth of a second. The
chamber must hold the droplet in flight for at least this long, and for the
slower falling-rate period that follows as the crust forms, so real spray-dryer
residence times run a few seconds with a margin. The $d^2$-law is why
atomisation is the whole game in spray drying: halving the droplet diameter
quarters the drying time, so the atomiser droplet-size distribution, not the
chamber size, sets whether the powder leaves dry, and a few oversized droplets
carry wet cores to the wall and cake it \cite{text:perry-chemical-engineers}.

\paragraph{Pressure-ratio limit in a single-stage gas membrane.} A gas
membrane cannot enrich the permeate beyond what the pressure ratio allows, no
matter how selective the membrane, and the ceiling is worth computing before a
membrane is chosen. We find the enrichment limit for a carbon-dioxide-selective
membrane separating $\text{CO}_2$ from methane in a biogas at feed mole
fraction $x_F = 0.40$, with a selectivity $\alpha = 40$ and a pressure ratio
$\varphi = p_f/p_p = 10$. The permeate mole fraction from the local flux
balance, in the limit of vanishing stage cut so the feed is undepleted,
solves a quadratic, but the two bounding limits are the ones to carry. The
selectivity-limited enrichment, reached at high pressure ratio, is
$y_P/(1 - y_P) = \alpha\,x_F/(1 - x_F) = 40(0.40/0.60) = 26.7$, giving
$y_P = 0.964$. The pressure-ratio-limited enrichment, reached at low pressure
ratio, caps the permeate partial pressure at the feed partial pressure:
$y_P \le \varphi\,x_F = 10(0.40) = 4.0$, which exceeds one and so does not
bind here, meaning the pressure ratio of ten is generous enough that the
membrane runs near its selectivity limit. Dropping the pressure ratio to
$\varphi = 2$ gives a binding cap $y_P \le 2(0.40) = 0.80$, well below the
selectivity ceiling of $0.964$, so at that pressure ratio the membrane wastes
its selectivity and the permeate cannot exceed $0.80$ however selective the
material. The rule the arithmetic fixes is that a high-selectivity membrane on
a low-pressure-ratio feed is pressure-ratio-limited and its selectivity is
squandered, which pushes gas-separation systems toward high feed pressure or a
vacuum permeate, and which is why the pressure ratio is checked against the
feed composition before the membrane selectivity is paid for
\cite{text:bird-transport-phenomena}.

\paragraph{Droplet evaporation time from the Ranz-Marshall correlation.} A
single falling droplet evaporating into a gas is the elementary event behind
spray drying, spray absorption, and combustion of a liquid fuel, and the
Ranz-Marshall correlation of section~7.2 turns the droplet's slip velocity into
a mass-transfer coefficient and thence a lifetime. We size the evaporation time
for a water droplet of diameter $d = 200\,\si{\micro\meter}$ falling through
still air at $25\,\degC$ and $50\,\%$ relative humidity at a terminal slip
velocity $u = 0.7\,\text{m/s}$, with air kinematic viscosity $\nu = 1.55\times
10^{-5}\,\si{\meter\squared\per\second}$ and water-vapour diffusivity
$D_{AB} = 2.6\times 10^{-5}\,\si{\meter\squared\per\second}$. The droplet
Reynolds number is $\text{Re} = ud/\nu = (0.7)(2.0\times 10^{-4})/(1.55\times
10^{-5}) = 9.0$, and the Schmidt number is $\text{Sc} = \nu/D_{AB} = 0.596$.
The Ranz-Marshall correlation gives
\[
\text{Sh} = 2 + 0.6\,\text{Re}^{1/2}\,\text{Sc}^{1/3} =
2 + 0.6\,(9.0)^{1/2}(0.596)^{1/3} = 2 + 0.6(3.0)(0.842) = 3.52.
\]
The floor of two is the pure-conduction limit of a stagnant sphere, and the
convective term adds half again here at this modest Reynolds number. The
coefficient is $k_c = \text{Sh}\,D_{AB}/d = (3.52)(2.6\times 10^{-5})/
(2.0\times 10^{-4}) = 0.46\,\text{m/s}$, high because the droplet is small and
the gas-phase diffusion path is short. The surface holds at the saturation
concentration $c_{\text{sat}}(25) = 3170/[(8.314)(298)] = 1.28\,\text{mol/m}^3$
against a bulk $0.50(1.28) = 0.64\,\text{mol/m}^3$, so the evaporative molar
flux is $N_A = k_c(c_{\text{sat}}-c_\infty) = (0.46)(0.64) = 0.29\,\text{mol}/
(\text{m}^2\cdot\text{s})$. The droplet holds
$n = (\pi/6)d^3\rho_L/M = (\pi/6)(2.0\times 10^{-4})^3(1000)/(0.018) =
2.3\times 10^{-7}\,\text{mol}$ of water over a surface $\pi d^2 =
1.26\times 10^{-7}\,\text{m}^2$, so the instantaneous loss is $N_A\pi d^2 =
3.7\times 10^{-8}\,\text{mol/s}$ and the crude lifetime is $n/(N_A\pi d^2) =
6.3\,\text{s}$. Because the surface area shrinks as the droplet evaporates, the
true lifetime follows the $d^2$-law and is shorter than this constant-area
estimate by a factor near three, landing a couple of seconds, which is why a
spray dryer must hold fine droplets in flight for only seconds and why a coarse
spray, with the square of the diameter against it, carries wet cores to the
wall.

\paragraph{Batch distillation by the Rayleigh equation.} A batch still has no
reflux and no operating line: the pot is charged, boiled, and the vapour drawn
off as it forms, so the pot composition drifts as the light component leaves
faster than the heavy one, and the Rayleigh equation, not a McCabe-Thiele
staircase, tracks it. We distil a charge of $W_0 = 100\,\text{mol}$ starting at
a light-component mole fraction $x_0 = 0.50$ with a constant relative volatility
$\alpha = 2.5$, and ask what the pot composition is after a third of the charge
has boiled over. The differential mole balance over the pot, light component
against total, is the Rayleigh relation
\[
\ln\!\frac{W}{W_0} = \int_{x_0}^{x}\frac{\mathrm{d}x}{y - x},
\]
with $y = \alpha x/[1+(\alpha-1)x]$ the vapour in equilibrium with the pot. For
constant relative volatility the integral has the closed form
\[
\ln\!\frac{W}{W_0} = \frac{1}{\alpha-1}\ln\!\frac{x}{x_0} +
\frac{\alpha}{\alpha-1}\ln\!\frac{1-x_0}{1-x},
\]
which we invert for $x$ at $W/W_0 = 2/3$, so the left side is $\ln(2/3) =
-0.405$. Trying $x = 0.41$: the first term is $(1/1.5)\ln(0.41/0.50) =
0.667(-0.199) = -0.133$, the second is $(2.5/1.5)\ln(0.50/0.59) =
1.667(-0.166) = -0.276$, summing to $-0.409$, within a percent of the target.
The pot has dropped from $0.50$ to about $0.41$ light component after a third of
the charge is gone, and figure~\ref{fig:vol04ch07:rayleigh-batch} reads the
same point off the lower-volatility curve. The average distillate composition
is the mass balance on the light component, $x_{D,\text{avg}} = (W_0 x_0 -
W x)/(W_0 - W) = (100\cdot 0.50 - 66.7\cdot 0.41)/33.3 = (50 - 27.3)/33.3 =
0.68$, richer than the pot as it must be. The lesson the arithmetic carries is
that a batch still trades sharpness for simplicity: without reflux it cannot
hold a product cut at constant composition, so it is run to a series of cuts,
and the Rayleigh integral is how the cut points are placed. Adding a rectifying
column and reflux over the pot recovers the McCabe-Thiele picture at the cost of
the hardware a batch operation was chosen to avoid.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-rayleigh-batch.tex}

\paragraph{Cooling-tower transfer units by the Merkel integral.} A cooling
tower rejects heat by evaporating a small fraction of the circulating water into
an air stream, and its sizing runs on a coupled heat-and-mass balance that the
Merkel analysis collapses into a single enthalpy driving force. We size the
transfer units for a tower cooling water from $38\,\degC$ to $28\,\degC$
against air entering at an enthalpy $h_1 = 72\,\text{kJ/kg}$ dry air, at a
water-to-air mass ratio $L/G = 1.2$. The heat given up by the water raises the
air enthalpy along an operating line of slope $L/G$, so the air leaves at
$h_2 = h_1 + (L/G)c_w(T_{w,\text{in}}-T_{w,\text{out}}) = 72 + 1.2(4.19)(10) =
72 + 50.3 = 122\,\text{kJ/kg}$. The Merkel number, the tower's transfer-unit
requirement, is the integral of the reciprocal enthalpy driving force over the
water temperature,
\[
\text{Me} = \frac{K a\,V}{L} = \int_{T_{\text{out}}}^{T_{\text{in}}}
\frac{c_w\,\mathrm{d}T_w}{h^{*}(T_w) - h(T_w)},
\]
with $h^{*}(T_w)$ the enthalpy of air saturated at the local water temperature
and $h(T_w)$ the air-operating-line enthalpy. Reading saturation enthalpies
$h^{*}$ at four water temperatures and the operating $h$ from the straight line,
the integrand $c_w/(h^{*}-h)$ is tabulated and integrated by Simpson's rule at
$T_w = 28, 31.3, 34.7, 38\,\degC$:
\[
\begin{array}{lcccc}
T_w\ (\si{\degreeCelsius}) & 28 & 31.3 & 34.7 & 38\\
h^{*}-h\ (\text{kJ/kg}) & 18 & 21 & 20 & 15\\
c_w/(h^{*}-h)\ (\text{K}^{-1}) & 0.233 & 0.200 & 0.210 & 0.279
\end{array}
\]
with panel width $\Delta T = 3.33\,\text{K}$, Simpson's rule gives
$\text{Me} = (\Delta T/3)[f_0 + 4f_1 + 2f_2 + \ldots]$ over the four points as
two panels, $\text{Me} = (3.33/3)[0.233 + 4(0.200) + 0.210] +
(3.33/3)[0.210 + 4(0.279)\cdot\tfrac12 \ldots]$, which to the accuracy of the
table lands near $\text{Me} \approx 1.6$. A Merkel number near $1.5$ to $2$ is
the working requirement for a ten-degree cooling range, and the packing height
follows from the tower's volumetric coefficient $Ka$ exactly as the packed
height followed from $H_{OG}$ in the scrubber. Figure~\ref{fig:vol04ch07:cooling-tower}
draws the operating line against the saturation curve and shows the driving
force pinching at the warm end, which is what sets the minimum air rate: tilting
the operating line up until it touches the saturation curve drives $\text{Me}$
to infinity, and a real tower is run with the air rate comfortably above that
minimum so the warm-end driving force stays open.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-cooling-tower.tex}

\paragraph{Nanofiltration and Donnan exclusion of a divalent salt.} A
nanofiltration membrane sits between reverse osmosis and ultrafiltration: it
passes monovalent salts fairly freely but rejects divalent ions and small
organics, and the mechanism that sorts them is not pore size alone but the
Donnan exclusion of co-ions by the membrane's fixed charge. We estimate the
rejection of magnesium sulphate by a negatively charged nanofiltration membrane
carrying a fixed charge density $X = 0.6\,\text{mol/L}$ against a feed of
$c_f = 0.02\,\text{mol/L}$ MgSO$_4$. Donnan equilibrium at the feed-membrane
interface balances the electrochemical potential of each ion, and for a
symmetric $z$:$z$ salt of valence $z$ the co-ion (here the divalent sulphate,
excluded by the negative fixed charge) partitions into the membrane at a
concentration set by
\[
\frac{c_{\text{co,mem}}}{c_{\text{co,feed}}} =
\left(\frac{c_{\text{co,feed}}}{c_{\text{co,mem}} + X/|z|}\right)^{1},
\]
which for a strongly charged membrane, $X \gg c_f$, reduces to the exclusion
scaling $c_{\text{co,mem}}/c_{\text{co,feed}} \approx (|z| c_f/X)^{1}$ raised to
a power that grows with valence. Putting the divalent salt into the working
estimate, the co-ion partition is roughly $(2 c_f/X) = 2(0.02)/0.6 = 0.067$ for
the divalent sulphate, so only about seven percent of the sulphate enters the
membrane and the salt rejection is near $1 - 0.067 = 0.93$, a $93\,\%$
rejection of the divalent salt at a fixed charge thirty times the feed
concentration. A monovalent salt at the same conditions partitions far less
strongly, $(c_f/X) = 0.033$ but with a valence exponent of one rather than the
steeper divalent dependence, and after the electroneutrality coupling of the
counter-ion its net rejection falls to perhaps $40$ to $60\,\%$, which is why
nanofiltration is chosen exactly where a stream must be softened, stripping
divalent hardness while letting monovalent salt through, at a fraction of the
pressure reverse osmosis would demand. The design reading is that
nanofiltration rejection is a charge phenomenon first and a size phenomenon
second, so it is sized against the ionic valences and the membrane charge, not
against a nominal pore diameter \cite{text:perry-chemical-engineers}.

\paragraph{Taylor-Aris dispersion in a laminar tube.} The axial-dispersion
coefficient that the packed-bed mastery box treats as a lumped input has a
first-principles value in the clean case of laminar flow through an open tube,
the Taylor-Aris result, and computing it shows why a slow flow disperses less
than intuition suggests. When a solute pulse travels down a tube in laminar
flow, the parabolic velocity profile shears it, spreading the pulse far faster
than molecular diffusion alone; but radial diffusion works against the shear,
carrying solute between the fast core and the slow wall, and the two effects
combine into an effective one-dimensional dispersion coefficient
\[
D_{\text{eff}} = D + \frac{u^2 R^2}{48\,D},
\]
with $u$ the mean velocity, $R$ the tube radius, and $D$ the molecular
diffusivity. We evaluate it for a solute of $D = 1.0\times 10^{-9}\,\si{
\meter\squared\per\second}$ in a tube of radius $R = 0.5\,\si{\milli\meter}$ at
a mean velocity $u = 0.02\,\text{m/s}$. The shear term is
\[
\frac{u^2 R^2}{48\,D} = \frac{(0.02)^2(5.0\times 10^{-4})^2}{48(1.0\times
10^{-9})} = \frac{(4.0\times 10^{-4})(2.5\times 10^{-7})}{4.8\times 10^{-8}} =
\frac{1.0\times 10^{-10}}{4.8\times 10^{-8}} = 2.1\times 10^{-3}\,\si{
\meter\squared\per\second},
\]
six orders of magnitude above the molecular $D$, so $D_{\text{eff}} \approx
2.1\times 10^{-3}\,\si{\meter\squared\per\second}$ and the shear dominates
utterly. The counter-intuitive feature is the inverse dependence on the
molecular diffusivity: a solute that diffuses radially faster is dispersed
axially less, because faster radial mixing evens the profile the shear is
trying to stretch. The result governs the band-broadening of a
chromatographic peak, the residence-time distribution of a tubular reactor, and
the smearing of a tracer pulse in a pipeline, and it is the open-tube parent of
the $D_{\text{ax}}$ that the packed bed carries as a correlation; a packed bed
adds the mechanical splitting and rejoining of streamlines to this laminar
core, which is why its particle Peclet number sits near two rather than at the
much larger value a bare tube would give \cite{text:bird-transport-phenomena}.

\paragraph{Minimum wash liquor for a counter-current leaching cascade.} A
counter-current washing train recovers a dissolved solute from a slurry of
spent solids, and the minimum wash-liquor rate below which the target recovery
cannot be met, no matter how many stages, is the leaching mirror of the minimum
reflux and the minimum solvent of the earlier operations. We size the minimum
wash for recovering $98\,\%$ of a dissolved sugar from a bed of spent beet cossettes
carrying $R = 0.4\,\text{kg}$ of retained liquor per kilogram of insoluble
solid, treating the solid as inert and the solute as fully dissolved in the
retained and wash liquors. The wash factor, the leaching analogue of the
stripping factor, is $W = V/(R\,\text{solid rate})$, the ratio of wash-liquor
flow to retained-liquor flow, and for a counter-current cascade with clean
entering wash the fraction of solute left with the solids after $N$ stages
follows the Kremser form
\[
f_{\text{lost}} = \frac{W - 1}{W^{N+1} - 1}.
\]
A recovery of $98\,\%$ leaves $f_{\text{lost}} = 0.02$, and the wash factor
must exceed one for any finite recovery, exactly as an absorption factor must.
At the infinite-stage limit the loss floor is $(W-1)/W^{N+1}\to 0$ only for
$W > 1$; at $W = 1$ the floor is $1/(N+1)$, so $98\,\%$ recovery at $W = 1$
would need $N + 1 = 1/0.02 = 50$ stages, absurd. Choosing a practical wash
factor $W = 1.5$, the stage count is $W^{N+1} = 1 + (W-1)/f_{\text{lost}} =
1 + 0.5/0.02 = 26$, so $N + 1 = \ln 26/\ln 1.5 = 3.258/0.405 = 8.0$ and
$N = 7$ washing stages. The wash liquor is then $V = W R\,(\text{solid rate}) =
1.5(0.4) = 0.60\,\text{kg}$ of wash per kilogram of solid, against an absolute
minimum near $W = 1$, or $0.40\,\text{kg/kg}$, below which no cascade reaches
the target. The engineering trade is the same one absorption and extraction
run: more wash liquor, a higher $W$, cuts the stage count but dilutes the
recovered solute and loads the downstream evaporator, so the wash rate is set a
modest margin above the minimum, and the recovered liquor is concentrated back
in a multiple-effect evaporator whose steam economy pays for the dilution the
washing chose \cite{text:treybal-mass-transfer}.

\section{Distillation: trays, packing, HETP, McCabe-Thiele}\pathtag{core}

Distillation separates a mixture by exploiting the difference in
volatilities (boiling points) of the components. Heating the mixture
produces a vapour richer in the more volatile component, which
condenses to a more concentrated liquid; condensing some of that
liquid and refluxing it back to the column repeats the enrichment.
The product at the top of the column approaches the more-volatile
pure component; the product at the bottom approaches the less-
volatile.

\engterm{Equilibrium-stage analysis}. The classical analysis models
the column as a sequence of stages, each at vapour-liquid equilibrium.
On each stage, the entering vapour from below mixes with the entering
liquid from above, the mixture reaches equilibrium, and the vapour
rises to the next stage while the liquid descends. The equilibrium
relation $y_A^* = f(x_A)$ for the system, the operating line
(material balance around the column), and the boundary conditions
(reflux at the top, reboiler at the bottom, feed composition and
location) determine the column's separation performance.

\engterm{McCabe-Thiele construction}. For binary distillation, the
McCabe-Thiele method gives a graphical solution on the
$y$-vs-$x$ diagram:

\begin{enumerate}[leftmargin=*]
  \item Plot the equilibrium curve $y^* = f(x)$ for the binary mixture.
  \item Plot the rectifying-section operating line, $y = (R/(R+1)) x
    + x_D/(R+1)$, where $R$ is the reflux ratio and $x_D$ the
    distillate composition.
  \item Plot the stripping-section operating line, $y = ((L+qF)/V')
    x - (W x_W - qF z_F)/V'$, where $W$ is the bottoms flow, $x_W$
    the bottoms composition, $F$ the feed flow, $z_F$ the feed
    composition, and $q$ the feed-thermal-state parameter.
  \item Step off stages alternately between the equilibrium curve
    and the operating lines, starting from the distillate composition,
    until the bottoms composition is reached. The number of steps
    is the number of equilibrium stages required for the separation.
\end{enumerate}

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-mccabe-thiele.tex}

A measured ethanol-water equilibrium table, the data behind the
azeotrope at $89$ mol-\% ethanol, is supplied in
\texttt{data/ethanol\_water\_vle.csv}; the stage-counting solver in
\texttt{code/mccabe\_thiele.py} steps the staircase of
figure~\ref{fig:vol04ch07:mccabe-thiele} for a constant relative
volatility and reports the stage count and feed location.

The minimum reflux ratio corresponds to the operating line touching
the equilibrium curve (an infinite number of stages required). The
total reflux corresponds to the operating line being the $y = x$
diagonal (the minimum number of stages, given by the Fenske equation).
Practical operation is at $1.2$ to $1.5$ times the minimum reflux;
the trade-off is between stage count (capital cost) and reflux
(operating cost).

\engterm{Tray and packed columns}. Two distillation hardware types
dominate:

\begin{itemize}[leftmargin=*]
  \item \engterm{Tray columns}: each tray is an actual physical
    plate where vapour bubbles up through a holdup of liquid. Trays
    approach but do not reach equilibrium; the \engterm{Murphree
    tray efficiency} $E_M$ (typically $0.5$ to $0.8$) accounts for
    the gap. The number of physical trays equals the number of
    equilibrium stages divided by $E_M$. Tray columns are robust
    against fouling and easy to clean.
  \item \engterm{Packed columns}: a continuous contact region filled
    with random or structured packing provides surface area for the
    vapour and liquid to interact. The separation is characterised by
    the \engterm{height equivalent to a theoretical plate} (HETP):
    the column height that produces one equilibrium stage of
    separation. Typical HETP values are $0.2$ to $0.6\,\text{m}$ for
    random packing and $0.1$ to $0.4\,\text{m}$ for structured
    packing. Packed columns offer lower pressure drop than tray
    columns and are preferred for vacuum-distillation services and
    for fragile feedstocks.
\end{itemize}

\engterm{Reactive distillation}. When the reaction is reversible and
fast, conducting the reaction within the distillation column
simultaneously drives the equilibrium toward products (by removing
them) and reduces the separate reactor requirement. Reactive
distillation columns appear in fuel-ether production (MTBE, ETBE),
in esterification reactions, and in metathesis processes. The
combined unit-operation design is more constrained than either
alone but can substantially reduce the capital cost.

\engterm{Multicomponent columns and the key components}. The McCabe-Thiele
construction is a binary tool, and most real distillations carry three or more
components, so the working method changes. A multicomponent column cannot make
every component pure at both ends at once from a single column; it splits the
feed between two components chosen as the split point, the \engterm{light key}
(the lightest component allowed to appear in appreciable amount in the bottoms)
and the \engterm{heavy key} (the heaviest allowed in the distillate). Components
lighter than the light key go almost entirely overhead and components heavier
than the heavy key go almost entirely to the bottoms; the separation is
specified by the recoveries of the two keys, and the components between them, if
any, distribute. The shortcut sizing runs the Fenske-Underwood-Gilliland chain:
the Fenske equation gives the minimum stages at total reflux from the key
recoveries and the key relative volatility, the Underwood equations give the
minimum reflux from the feed composition and volatilities through a root that
lies between the two key volatilities, and the Gilliland correlation maps the
actual reflux onto the actual stage count between those two minima, the same
correlation the binary exercises use. The chain is a first cut that sizes the
column before a stage-by-stage or rate-based simulation refines it, and it is
the multicomponent analogue of the binary Fenske-plus-Gilliland arithmetic of
the estimation block.

\engterm{Azeotropic and extractive distillation}. When the relative volatility
falls to one at an azeotrope, or is simply too close to one over the whole
composition range, ordinary distillation stalls, and the fix is to change the
volatility ratio by adding a third component. \engterm{Extractive distillation}
adds a high-boiling solvent that stays in the liquid throughout the column and
selectively lowers the volatility of one feed component, opening the relative
volatility of the pair and letting the column separate them; the solvent leaves
in the bottoms and is recovered in a second column. The dehydration of ethanol
with ethylene glycol as the extractive solvent is the classic case, pulling
water preferentially into the liquid so ethanol distils overhead past the
azeotrope. \engterm{Azeotropic distillation} instead adds a low-boiling
entrainer that forms a new azeotrope with one component, carrying it overhead in
a way that breaks the original azeotrope; the entrainer is recovered from the
overhead by decantation or a further column. Both routes pay for crossing the
azeotrope with a third component and its recovery train, which is the cost the
pervaporation vignette avoids by separating on a membrane's permeability ratio
rather than on the vapour-liquid equilibrium the entrainer has to bend.

\begin{mastery}
The equilibrium-stage model with a Murphree efficiency is a fiction that works
until it does not. It pretends each tray reaches vapour-liquid equilibrium and
then patches the gap with a single fudge factor, the efficiency, which is
assumed the same for every component and every tray. For a binary at modest
purity the fiction is harmless; for a multicomponent column, a highly
non-ideal mixture, or a reactive service, it fails, because the Murphree
efficiency is not a property of the tray but of the tray, the mixture, and the
operating point together, and it can differ between components on the same
tray, even exceeding one or going negative for the species being dragged by
the others in the Maxwell-Stefan sense of the first section. The rate-based, or
nonequilibrium, model replaces the equilibrium-plus-efficiency assumption with
an explicit account of the transport. Each tray or packing segment carries a
mass balance for each phase, an energy balance for each phase, and an
interface where the two phases are in equilibrium only at the contact surface,
\emph{not} in the bulk. The interphase fluxes are written from the two-film or
Maxwell-Stefan relations of this chapter, the interfacial area comes from the
tray or packing hydraulics, and the model solves the coupled bulk balances and
interface equilibria simultaneously. Nothing in it is a theoretical stage: the
column is a stack of real trays or a length of real packing, and the
separation falls out of the transport rates and the interfacial area rather
than out of a stage count times an efficiency. The payoff is that the
rate-based model predicts the composition and temperature profiles of a
difficult column, a reactive column, or a column with large heat effects, that
the equilibrium model gets wrong; the cost is that it demands the binary
diffusivities, the tray or packing mass-transfer correlations, and the
interfacial-area data, and it solves a far larger set of coupled equations.
The working judgment is to reach for the equilibrium-stage model with an
efficiency for a routine binary or near-ideal separation, and for the
rate-based model when the mixture is strongly non-ideal, multicomponent with a
dragging species, reactive, or heat-effect dominated, where the single-number
efficiency has no honest value to take.
\end{mastery}

\section{Absorption, extraction, adsorption}\pathtag{standard}

Three separation operations transfer species between phases:

\engterm{Absorption}: a gas mixture is contacted with a liquid that
selectively dissolves one component. Examples: $\text{CO}_2$
absorption from natural gas by an amine solution; $\text{SO}_2$
removal from flue gas by aqueous limestone slurry; ammonia
absorption from a synthesis-loop purge by water; hydrogen sulphide
absorption from sour gas by methyl-diethanolamine.

The absorber design follows the equilibrium-stage analysis applied
to a counter-current gas-liquid column: gas rises, liquid descends,
absorption transfers solute from gas to liquid. The Kremser equation
gives the closed-form solution for absorption with linear equilibrium
($y_A^* = m x_A$):
\[
\frac{Y_{A,N+1} - Y_{A,1}}{Y_{A,N+1} - m X_{A,0}} = \frac{A^{N+1} - A}{
A^{N+1} - 1},
\]
where $A = L/(mG)$ is the absorption factor, $N$ the number of
stages, and $Y_A, X_A$ are solute-to-carrier mole ratios. For
$A > 1$, the absorber is well-behaved and increases stage count
improve recovery; for $A < 1$, the absorber cannot recover all the
solute even with infinite stages. The closed-form Kremser sizing,
including the inversion for the stage count at a target recovery, is
implemented in \texttt{code/kremser\_absorber.py}.

\engterm{Liquid-liquid extraction}: two immiscible liquids are
contacted, with the solute distributing between them according to
its partition coefficient $K = y_A/x_A$. Extraction is used to
recover dilute species from a feed liquid (penicillin from
fermentation broth, lithium from brine, hydrocarbons from waste
streams) and to separate pairs of liquids of similar volatility
that distillation cannot economically separate. The stage analysis
mirrors absorption with the extraction factor $E = K\,(S/F)$ playing
the role of the absorption factor: $E > 1$ makes the raffinate
recoverable to any target with enough stages, while $E < 1$ leaves a
floor of unrecovered solute. The choice of solvent turns on three
numbers at once: a high partition coefficient (so few stages suffice),
a low mutual solubility with the feed (so the solvent is not lost into
the raffinate), and a density difference and interfacial tension that
let the phases disengage in the settler. A solvent with a superb
partition coefficient that emulsifies with the feed is worthless,
because the settler cannot separate the phases; extraction equipment
selection (mixer-settlers, packed or pulsed columns, centrifugal
contactors) is driven as much by the ease of phase disengagement as by
the stage requirement.

The single partition coefficient hides a geometry that the ternary phase
diagram makes plain, and the geometry is what governs a real extraction where
the solvent and the feed are not perfectly immiscible. Plotted on a triangular
diagram whose corners are the solute, the feed solvent, and the extracting
solvent, the two-liquid region is a dome under a binodal curve, and any mixture
inside the dome splits into two conjugate phases whose compositions are the two
ends of the tie line through the mixture point. The slope of the tie lines is
the graphical statement of the partition coefficient: a steep tie line sends
most of the solute into the extract phase and makes for an easy separation,
while tie lines that flatten toward the plait point, where the two phases
become identical and the partition coefficient goes to one, mark the
composition beyond which no extraction is possible. A single equilibrium stage
is one tie line: the feed and solvent mix to a point inside the dome, and the
two phases that settle out are read off the ends of the tie line through it. A
counter-current cascade steps between the binodal and an operating line the way
McCabe-Thiele steps between the equilibrium curve and its operating line, with
the number of tie-line steps the stage requirement. Two features that
distillation does not carry appear here. The extracting solvent dissolves into
the raffinate and is lost with it, so a solvent-recovery step, often a
distillation, hangs off the back of every extraction and its cost enters the
solvent choice. And the mutual solubility grows as the operating point nears
the plait point, so an extraction pushed toward high recovery runs into a
composition where the phases refuse to separate cleanly, the geometric root of
the phase-disengagement problem the solvent choice must avoid. The partition
coefficient is the first number, but the shape of the dome and the slope of the
tie lines are what a real ternary extraction is sized against.

\engterm{Adsorption}: a solid sorbent selectively binds species from
a gas or liquid phase. Examples: activated-carbon removal of
volatile organic compounds from water and air; molecular-sieve
drying of natural gas to pipeline specifications; pressure-swing
adsorption to separate oxygen from nitrogen in air for medical and
industrial use. Adsorption is operated batchwise (the bed loads
with solute until breakthrough, then is regenerated by heat,
pressure, or a wash) or in moving-bed configurations (rotary
adsorbers, simulated moving beds).

The adsorption design depends on the adsorbent's isotherm (the
loading of solute on the solid as a function of fluid-phase
concentration). For linear isotherms (Henry's law range), the design
parallels absorption; for non-linear (Langmuir, Freundlich, BET)
isotherms, the design requires breakthrough-curve analysis. The shape
of the isotherm controls how the mass-transfer zone travels down the
bed. A favourable (concave-down, Langmuir-type) isotherm sharpens the
concentration front into a near-constant-pattern wave, so the
breakthrough is steep and the bed is used almost to its equilibrium
capacity before the outlet rises. An unfavourable (concave-up) isotherm
spreads the front, broadening the mass-transfer zone, wasting capacity,
and forcing early switchover. The working sizing number is the
length of unused bed, the difference between the bed length and the
length that would be fully loaded at breakthrough; a sharp front leaves
little unused bed, a broad front leaves much. The breakthrough trace of
figure~\ref{fig:vol04ch07:breakthrough} is the measurement that fixes
that number for a given sorbent, feed, and flow rate.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-breakthrough.tex}

A measured breakthrough trace for a fixed-bed adsorber is supplied in
\texttt{data/adsorption\_breakthrough.csv}; integrating it against the
feed concentration gives the bed's working capacity and the fraction of
the equilibrium loading actually used before breakthrough.

The three isotherm models the design reaches for are worth writing down
plainly, because each carries a different assumption about the surface and
each is the right fit in a different regime. The \engterm{Langmuir isotherm}
assumes a fixed number of identical sites, each holding at most one molecule
with no interaction between neighbours, and gives
\[
q = \frac{q_m\,b\,c}{1 + b\,c},
\]
with $q_m$ the monolayer saturation loading and $b$ the affinity constant. At
low concentration it is linear, $q \approx q_m b\,c$, the Henry range where
adsorption parallels absorption; at high concentration it saturates at $q_m$
as the surface fills. Its concave-down shape is favourable, which is why the
nitrogen isotherm of the oxygen-concentrator case study is Langmuir and
sharpens its front. The \engterm{Freundlich isotherm},
\[
q = K_F\,c^{1/n} \quad (n > 1),
\]
is an empirical power law that fits heterogeneous surfaces with a spread of
site energies, and it is the workhorse for activated carbon loading organics
from water, where no single site energy describes the carbon. It has no
saturation, so it must not be extrapolated past the data. The \engterm{BET
isotherm} extends Langmuir to multilayer adsorption, admitting a second and
further layers on top of the first, and is the model behind surface-area
measurement by nitrogen adsorption at cryogenic temperature. The single
number that sorts the front behaviour is the dimensionless separation
constant $R_s = 1/(1 + b\,c_0)$ read off the Langmuir form at the feed
concentration $c_0$: $R_s < 1$ is favourable and self-sharpens the front,
$R_s = 1$ is linear, and $R_s > 1$ (an unfavourable isotherm) spreads it.
Figure~\ref{fig:vol04ch07:isotherm-shapes} draws the three shapes on a common
axis and marks how the favourable curve catches its own front while the
unfavourable one lets it disperse.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-isotherm-shapes.tex}

\begin{mastery}
The equilibrium-stage and plug-flow pictures both assume the fluid marches
through the bed or column in a flat front, every element spending the same
time in contact. Real packed beds and columns depart from plug flow: some
fluid channels ahead through the loose spots and some lags in the slow ones,
so the front smears even when the isotherm or the equilibrium line would keep
it sharp. Axial dispersion is the name for that smearing, and the
dimensionless group that measures it is the bed Peclet number,
\[
\mathrm{Pe} = \frac{u\,L}{D_{\text{ax}}},
\]
with $u$ the interstitial velocity, $L$ the bed length, and $D_{\text{ax}}$ an
effective axial-dispersion coefficient that lumps molecular diffusion,
Taylor-Aris dispersion in the pore channels, and the mechanical mixing of the
splitting and rejoining streamlines around the packing. A large Peclet number
is near-plug-flow, where convection dominates the axial mixing and the front
stays sharp; a small Peclet number is near-mixed, where the dispersion broadens
the front over a length comparable to the bed and steals separation. For
random packing a working correlation puts the particle Peclet number
$u\,d_p/D_{\text{ax}}$ near two in liquid service and near unity to two in gas,
so the bed Peclet number is roughly $2L/d_p$, which is why a bed must be many
particle diameters long to behave as plug flow: a short bed of coarse packing
disperses badly. The number of theoretical stages a real column delivers is
capped by its axial dispersion through a relation like $N \approx \mathrm{Pe}/2$
in the dispersed-plug-flow limit, so a column that channels or is packed too
short cannot reach its equilibrium-stage count no matter how good the mass
transfer on each element. The design consequences are direct: size the bed
long enough that $\mathrm{Pe}$ is large, distribute the feed evenly so the
dispersion is not compounded by gross maldistribution, and, when a
laboratory column reports fewer stages than the mass transfer predicts,
suspect axial dispersion before suspecting the equilibrium data. Axial
dispersion is the continuous-contact cousin of the tray efficiency: both are
the gap between the idealised stage count and the one the hardware actually
delivers.
\end{mastery}

\begin{mastery}
An adsorption bed carries two waves down its length, not one, and the
second wave is thermal. Adsorption releases heat, the heat of adsorption
$\Delta H_a$, of order $10\,\text{to}\,50\,\text{kJ/mol}$ for physical
adsorption and more for chemisorption. As the concentration front loads
the solid, that heat warms the gas and the sorbent, and a temperature wave
travels with or behind the concentration wave depending on the relative
speeds. The two wave speeds follow from the two conservation laws. The
concentration wave moves at the speed set by the isotherm slope,
\[
u_c = \frac{u}{1 + \dfrac{1-\varepsilon}{\varepsilon}\,\rho_p\,
\dfrac{\mathrm{d}q}{\mathrm{d}c}},
\]
with $u$ the interstitial gas velocity, $\varepsilon$ the bed voidage,
$\rho_p$ the particle density, and $\mathrm{d}q/\mathrm{d}c$ the local
isotherm slope; a steep isotherm slope (strong adsorption) slows the
front. The thermal wave moves at the analogous speed set by the heat
capacities,
\[
u_T = \frac{u\,\rho_g c_{p,g}}{\rho_g c_{p,g} + \dfrac{1-\varepsilon}
{\varepsilon}\rho_p c_{p,s}},
\]
the ratio of the gas heat capacity flowing in to the total heat capacity
that must be warmed. When the two waves travel at the same speed they ride
together and the temperature rise sits right on the mass-transfer zone,
which lowers the local capacity (a hotter sorbent holds less, by the
isotherm's temperature dependence) and can broaden the front into a
self-sharpening-then-spreading shape that no isothermal analysis predicts.
When the thermal wave outruns the concentration wave, the heat is carried
ahead and preheats the unloaded bed, raising the apparent equilibrium the
incoming front must overcome. The design consequence is that an isothermal
breakthrough calculation overestimates the working capacity whenever the
heat of adsorption is appreciable and the bed is large enough to be
adiabatic, which is most industrial beds; the cure is either to size the
bed against the measured nonisothermal breakthrough, to add internal
cooling, or, in a fast pressure-swing cycle, to let the small bed and the
short cycle keep the temperature excursion bounded. The oxygen
concentrator of the third case study is small and fast enough that the
thermal swing is modest, which is part of why a soda-can of sieve works at
all; a large adiabatic guard bed would carry a temperature wave tens of
degrees high and would lose a fifth or more of its isothermal capacity to
it.
\end{mastery}

\section{Membrane processes: RO, UF, MF, gas separation}\pathtag{standard}

A membrane is a thin barrier with selectivity: it admits some
species and rejects others. Membrane separations are driven by a
pressure difference (reverse osmosis, ultrafiltration,
microfiltration, nanofiltration), a concentration difference
(dialysis), an electrical potential difference (electrodialysis), or
a partial-pressure difference (gas separation). The membrane's
performance is characterised by two numbers: the flux of permeate
and the rejection (or selectivity) for the retained species.

\engterm{Reverse osmosis} (RO): a pressure greater than the osmotic
pressure of the feed is applied to drive water through the membrane
while rejecting dissolved salts. RO is the working desalination
technology, accounting for the bulk of new desalination capacity
since 2000 and supplying drinking water to several hundred million
people worldwide. The osmotic pressure of seawater is $\sim 2.7\,\text{MPa}$;
working RO systems operate at $\sim 6$ to $8\,\text{MPa}$ feed pressure
to provide adequate driving force across the membrane. The specific
energy consumption of seawater desalination has fallen from $\sim 8\,
\text{kWh/m}^3$ in the 1980s to $\sim 3\,\text{kWh/m}^3$ in the 2020s,
within a factor of $2$ of the thermodynamic minimum
\cite{gen:mackay2009}.

\engterm{Ultrafiltration} (UF): a lower-pressure process (typically
$0.1$ to $1\,\text{MPa}$) used to separate macromolecules from water
or small solutes. Applications include cheese whey processing,
biopharmaceutical concentration, and water-treatment polishing.

\engterm{Microfiltration} (MF): the lowest-pressure pressure-driven
membrane operation, used to remove bacteria, suspended solids, and
turbidity from water. The driving pressure is typically $\sim
0.1\,\text{MPa}$.

\engterm{Nanofiltration} (NF) sits between reverse osmosis and
ultrafiltration, run at $0.5$ to $1.5\,\text{MPa}$, and it separates by a
mechanism the others do not lean on: the fixed charge on the membrane repels
co-ions of the same sign, so a negatively charged nanofiltration membrane
rejects divalent anions such as sulphate strongly while passing monovalent
chloride fairly freely, the Donnan exclusion the nanofiltration worked example
sizes. The application is water softening, stripping divalent hardness while
leaving monovalent salt, at a fraction of the pressure reverse osmosis would
demand, and the design rule is that nanofiltration rejection is a charge
phenomenon first and a size phenomenon second.

The four pressure-driven operations sort by what sets the rejection, and the
distinction is the working reason each is chosen for its own duty. Microfiltration
and ultrafiltration are \engterm{pore-flow} membranes: the water moves through
open pores by viscous flow, and the membrane rejects by size, straining out
anything larger than the pore, so the flux is high and the rejection is sharp at
the pore cutoff. Reverse osmosis is a \engterm{solution-diffusion} membrane: the
water dissolves into a dense, effectively non-porous polymer film, diffuses
across down its own activity gradient, and desorbs on the permeate side, while
the salt is rejected because it dissolves in and diffuses through the film far
less readily than water; there is no pore for the salt to be strained by, and
the rejection is a ratio of diffusion-solubility products, not a size cutoff.
Nanofiltration straddles the two, with both a loose pore structure and a fixed
charge, which is why its rejection cannot be read off a pore size alone. The
consequence for design is that a pore-flow membrane is sized against the
particle or molecule size it must strain and fails by fouling that plugs the
pores, while a solution-diffusion membrane is sized against the osmotic
back-pressure and the permeability, and its rejection is set by the polymer
chemistry rather than by a hole. The same solution-diffusion picture carries
over to the dense gas-separation membranes below, where the permeating species
dissolves into the polymer and diffuses across exactly as water does in reverse
osmosis, and the selectivity is again a ratio of diffusion-solubility products.

\engterm{Gas separation} membranes select among gas species by
differences in their permeabilities through the membrane material.
Working applications: nitrogen generation from air (using oxygen-
selective polymer membranes), hydrogen recovery from synthesis-loop
purges (using palladium-alloy membranes), carbon-dioxide separation
from natural gas. The selectivity-permeability trade-off (the
``Robeson upper bound'') is a recurring constraint: more selective
membranes have lower permeability and thus require more membrane
area for a given throughput. Representative pressure, flux, and
selectivity figures for the pressure-driven and gas-separation
processes are collected in \texttt{data/membrane\_performance.csv}.

The working model for a dense polymer membrane is solution-diffusion: a
gas dissolves into the upstream face, diffuses across, and desorbs on
the permeate side, so the permeability is the product of a solubility
and a diffusivity. The flux of a species is its permeance times the
partial-pressure difference across the membrane, $N_i = \Pi_i (p_{f,i} -
p_{p,i})$, and the separation between two species is governed by the
ratio of their permeances, the selectivity $\alpha_{ij} = \Pi_i/\Pi_j$.
Two features of this flux law shape every gas-membrane design. The
driving force is a partial-pressure difference, not a total-pressure
difference, so enriching the permeate raises its partial pressure and
shrinks its own driving force; the purest permeate is the one taken in
the thinnest slice off the feed. And the pressure ratio
$p_f/p_p$ caps the achievable enrichment independently of selectivity:
a permeate cannot be more enriched than the feed by more than the
pressure ratio allows, because beyond that the permeate partial
pressure would exceed the feed partial pressure and the flux would
reverse. A high-selectivity membrane on a feed at low pressure ratio is
pressure-ratio-limited and wastes its selectivity, a trap that pushes
gas-separation systems toward high feed pressure or vacuum permeate.

\begin{mastery}
The Robeson upper bound is an empirical line on a log-log plot of
selectivity against permeability: across decades of polymer-membrane
data, the most selective materials are the least permeable, and no
polymer crosses the bound. Plotting $\alpha_{ij}$ (the ratio of the two
gas permeabilities) against the permeability $P_i$ of the faster gas,
the achievable region sits below a line of negative slope. The bound is
a constraint, not a law of nature: it reflects the diffusion-solubility
trade-off in dense polymers, where a stiffer, more size-selective chain
admits gas more slowly. Materials that beat the polymer bound, such as
carbon molecular sieves, mixed-matrix membranes, and metal-organic
frameworks, do so by separating on a different mechanism (molecular
sieving rather than solution-diffusion). The design lesson is that for
a polymer membrane the engineer cannot have both high selectivity and
high flux, and must size the module against whichever the separation
demands.
\end{mastery}

\engterm{Membrane fouling and concentration polarisation}. The two
operational difficulties of membrane processes are fouling (solids
or biological growth accumulating on the membrane surface) and
concentration polarisation (the rejected species building up at the
membrane surface, raising the local osmotic pressure and reducing
the driving force). Both are mitigated by cross-flow operation (the
feed flows tangentially across the membrane, sweeping the
accumulated solute away), by pretreatment (removing fouling
precursors from the feed), and by periodic cleaning.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-membrane-flux.tex}

\begin{mastery}
Concentration polarisation has a closed form that every membrane
designer carries. In a pressure-driven membrane the permeate water
leaves the rejected solute behind, so solute is swept toward the
membrane wall at the permeation velocity $J_w$ and diffuses back toward
the bulk down its own gradient. At steady state in the boundary-layer
film, a solute balance with permeate carrying a wall concentration
$c_w$ and a fully rejecting membrane reads
\[
J_w\,c - D\,\frac{dc}{dy} = J_w\,c_p,
\]
with $c_p$ the permeate concentration. Integrating across the film of
thickness $\delta$ from the bulk $c_b$ to the wall $c_w$, with the
liquid-side coefficient $k = D/\delta$ and a fully rejecting membrane
($c_p = 0$), gives the polarisation factor
\[
\frac{c_w}{c_b} = \exp\!\Big(\frac{J_w}{k}\Big).
\]
The wall concentration rises exponentially in the ratio of the
permeation velocity to the mass-transfer coefficient. The engineering
consequences are immediate. The osmotic pressure that the applied
pressure must beat is the wall value, $\pi(c_w)$, not the bulk value, so
polarisation steals driving force exactly where the membrane is working
hardest, and a unit sized on the bulk osmotic pressure underdelivers
flux. Worse, the exponential is self-limiting in the wrong direction: a
higher flux raises the wall concentration, which raises the osmotic
back-pressure, which suppresses the flux, so the operating point settles
below the naive prediction. The cure is to raise $k$ by increasing the
cross-flow velocity, narrowing the feed channel, or roughening the
spacer to promote mixing, all of which thin the film $\delta$ and pull
the exponent back toward zero. For a gas-separation membrane the same
analysis applies to the slower component building up on the feed side,
and for ultrafiltration of macromolecules the wall concentration can
reach the gel point, capping the flux independently of pressure, the
phenomenon of pressure-independent limiting flux that no amount of added
pressure can beat.
\end{mastery}

\section{Drying, crystallisation, leaching}\pathtag{standard}

\engterm{Drying} removes a liquid (typically water) from a solid by
vapourisation. The design problem is the rate of moisture removal as
a function of solid moisture content, the air conditions, and the
dryer geometry. The drying curve has two regimes:

\begin{itemize}[leftmargin=*]
  \item Constant-rate period: while the solid surface is saturated
    with liquid, the drying rate is set by the air-side mass-transfer
    coefficient and the saturation deficit (the difference between
    saturation vapour pressure of the liquid and the partial pressure
    in the bulk air). The surface temperature equals the wet-bulb
    temperature.
  \item Falling-rate period: as the solid moisture content falls
    below the critical value, the surface dries faster than liquid
    can be transported from the interior; the rate falls and the
    solid temperature rises toward the dry-bulb air temperature.
\end{itemize}

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-drying-curve.tex}

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-evaporation.tex}

The two periods rest on two different controlling resistances, and knowing
which one governs is what lets a dryer be sized rather than guessed. In the
constant-rate period the solid surface runs wet, the resistance sits entirely
in the gas-side boundary layer, and the rate is the external mass-transfer
flux $k_c(c_{\text{sat}} - c_\infty)$ over the exposed area, the same
convective transport the wetted-wall and pool examples computed. Because the
evaporating surface holds at the wet-bulb temperature, the constant-rate flux
also follows from a heat balance, $h(T_\infty - T_w)/(M_A h_{fg})$, and the two
routes agree through the near-unity Lewis number of the air-water system that
the wet-bulb example used. Raising the air velocity, lowering its humidity, or
raising its temperature all speed the constant-rate period, because all three
act on the external boundary layer. Once the moisture falls below the critical
value the surface can no longer stay wet, the wet patches shrink and the dry
patches spread, and the controlling resistance moves inside the solid. In the
falling-rate period the rate is set by how fast moisture reaches the surface
from the interior, which is a diffusion problem in the solid with a diffusivity
that is itself a strong function of moisture content and structure. For a solid
whose internal transport is Fickian, the falling-rate flux is roughly
proportional to the free moisture $W - W^{*}$, which gives the exponential
approach to the equilibrium moisture that the drying-time worked example
integrated; for a solid with a receding wet front the rate follows a
shrinking-core law and the time scales as the square of the piece size, the
same $L^2/D$ signature as any internal-diffusion limit. The two-period split is
why speeding the air helps a wet solid and does almost nothing for a nearly dry
one: the external boundary layer governs the first period and the internal
diffusion governs the second, and only shrinking the piece, raising its
temperature, or opening its structure touches the internal resistance. Sizing a
batch dryer on the constant-rate period alone, as the worked example warned,
underpredicts the cycle by the whole of the falling-rate contribution, which
for a thick or dense solid is the larger share.

Industrial dryers include rotary dryers, fluidised-bed dryers, spray
dryers (the workhorse of milk powder and instant-coffee production),
drum dryers (for syrups, emulsions), and freeze dryers (for
pharmaceuticals, instant coffee, food).

\engterm{Crystallisation} produces a solid product from a saturated
solution by reducing the solubility (cooling, evaporating, or
adding a non-solvent). The design targets are the crystal-size
distribution, the crystal habit (shape), and the purity, and the three
trade against one another through the supersaturation, the single driving
force of the operation. Supersaturation is the excess of the dissolved
concentration over the solubility, and it splits its work between two
competing kinetics: nucleation, which makes new crystals, and growth,
which enlarges the ones already present. High supersaturation favours
nucleation, producing many small crystals and a fine, hard-to-filter
product; low supersaturation favours growth, producing fewer, larger,
cleaner crystals that filter and wash well. The crystalliser's job is to
hold the supersaturation in the metastable zone, above the solubility so
crystals grow but below the spontaneous-nucleation threshold so the
solution does not flood with fines, often by seeding with a controlled
mass of small crystals and then ramping the cooling or evaporation slowly
enough that growth onto the seeds soaks up the supersaturation as it is
generated. The population-balance equation tracks the crystal-size
distribution as a conservation law in size space, with nucleation as a
source at zero size and growth as convection up the size axis; coupled to
the nucleation and growth rate laws and a solute balance, it predicts the
product distribution from the operating policy. Industrial
crystallisation produces salts, sugars, pharmaceutical actives, and many
fine chemicals, and in pharmaceutical practice it also controls which
crystal polymorph forms, a choice that sets the drug's dissolution rate
and shelf stability.

The two competing kinetics can be written down and their balance made
quantitative, which turns the operating-policy choice from art into a sizing
calculation. The supersaturation is the driving force, most usefully written
as the relative supersaturation $\sigma = (c - c^{*})/c^{*}$, the fractional
excess of the dissolved concentration over the solubility $c^{*}$. Crystal
growth is diffusion-and-integration limited and its linear growth rate
follows a power law in the supersaturation,
\[
G = \frac{\mathrm{d}L}{\mathrm{d}t} = k_g\,\sigma^{g},
\]
with the growth order $g$ near one to two for most systems. Nucleation, the
birth of new crystals, follows a much steeper power law,
\[
B = k_b\,\sigma^{b},
\]
with the nucleation order $b$ commonly four to ten, so a small rise in
supersaturation multiplies nucleation far faster than growth. The ratio of
the two, $B/G \propto \sigma^{b-g}$, climbs steeply with supersaturation, and
that steepness is the whole reason the crystalliser must hold $\sigma$ low: at
high $\sigma$ nucleation swamps growth and the product is a cloud of fines
that will not filter, while at low $\sigma$ growth dominates and the seeds
already present soak up the excess into fewer, larger, cleaner crystals. The
metastable zone is the band of supersaturation above the solubility, where
crystals grow, but below the spontaneous-nucleation threshold, where the
solution floods with new crystals; the operating policy is to seed with a
known mass of small crystals and then ramp the cooling or evaporation slowly
enough that the growth onto the seeds consumes the supersaturation as fast as
the cooling generates it, keeping $\sigma$ inside the metastable band the
whole way.

\engterm{Leaching} (solid-liquid extraction) transfers a solute
from a solid to a contacting liquid. Examples: coffee brewing
(leaching of soluble coffee components from ground beans), copper
heap leaching (leaching of copper from low-grade ore by sulphuric
acid percolation), and pharmaceutical extraction from plant material.
Three resistances act in series, and the slowest sets the rate: the
diffusion of solvent into the solid, the dissolution of the solute at
the internal surface, and the diffusion of dissolved solute back out
through the solid pores to the bulk liquid. For a coarse, low-porosity
solid the internal diffusion path dominates and the leach time scales as
the particle size squared, the same $L^2/D$ scaling as transient
diffusion, which is why grinding the ore or the bean is the first lever:
halving the particle size quarters the diffusion time. The contact
pattern then sets how completely the solute is recovered. A single batch
soak leaves the solid in equilibrium with a partly loaded liquid and
recovers only a fraction; a counter-current multi-stage cascade, with
fresh solvent meeting the most-depleted solid and loaded liquid leaving
against the fresh feed solid, drives the recovery toward completion with
the same Kremser-style stage analysis that governs absorption and
extraction, the partition coefficient between solid and liquid playing
the role of the equilibrium slope. Heap leaching of copper is the
continuous-percolation limit of the same picture, run over months rather
than minutes because the ore cannot be ground to a powder economically
and the internal diffusion path is correspondingly long.

The internal-diffusion limit that dominates a coarse solid has a clean model in
the shrinking-core picture, and it explains the leach-time scaling that grinding
exploits. Take a spherical particle of radius $R_0$ whose solute leaches out
from the surface inward, leaving a spent, porous shell through which fresh
solute must diffuse to reach the still-loaded core. If the diffusion through the
shell is the slowest step, the unreacted-core radius $r_c$ shrinks according to
a balance between the diffusive flux through the shell and the rate at which the
core is stripped, and integrating that balance gives the time to reach a
fractional recovery $X = 1 - (r_c/R_0)^3$,
\[
t = \frac{\rho_s R_0^2}{6\,D_e\,c_s}\left[1 - 3(1 - X)^{2/3} + 2(1 - X)\right],
\]
with $\rho_s$ the solute loading per unit particle volume, $D_e$ the effective
diffusivity through the spent shell, and $c_s$ the bulk solvent concentration
of the leaching agent. The bracket runs from zero at $X = 0$ to one at complete
leaching, so the time to a given recovery scales as $R_0^2/D_e$, the square of
the particle size, which is the quantitative form of the rule that halving the
particle size quarters the leach time. For a fine, fast leach the surface
reaction or the film may govern instead, in which case the time scales only
linearly with $R_0$, but a coarse ore or bean sits in the diffusion-controlled
regime where the square law rules and the economic question is how far it pays
to grind: each halving of the particle quarters the time but multiplies the
grinding energy, and the crossover sets the particle size a plant actually runs.
Heap leaching accepts the long diffusion time precisely because grinding the ore
to escape the square law would cost more than the months of percolation save.

\section{An end-to-end case study: a packed-column flue-gas scrubber}\pathtag{standard}

The chapter's pieces come together when an engineer must size a real
column from stream data alone. We take a packed-column absorber that
scrubs sulphur dioxide from a boiler flue gas with a dilute caustic
solution, and carry it from the inlet streams through the equilibrium
and operating lines, the number and height of transfer units, the
packed height, the column diameter from flooding, the irrigated
pressure drop, and a verdict an engineer would sign. The scrubber is a
good vehicle because every working move of the chapter appears in it
once: a gas-liquid equilibrium, a counter-current material balance, the
transfer-unit decomposition of section~7.3 and the absorption analysis
of section~7.5, the flooding correlation that fixes the diameter, and a
pressure-drop check that closes the design.

\subsection{The streams and the duty}

A coal-fired industrial boiler discharges $G = 12\,\text{m}^3/\text{s}$
of flue gas at $1\,\text{atm}$ and $50\,\degC$, carrying a sulphur
dioxide mole fraction $y_{\text{in}} = 0.0020$ ($2000\,\text{ppm}$).
The permit limit at the stack is $200\,\text{ppm}$, so the scrubber must
achieve $90\,\%$ removal: $y_{\text{out}} = 2.0\times 10^{-4}$. The
solvent is a dilute sodium-hydroxide solution that reacts with the
dissolved \text{SO}$_2$; for sizing we treat the reaction as fast enough
to keep the liquid-side equilibrium partial pressure low, so the working
equilibrium slope in mole-ratio coordinates is $m = 0.86$, a modest
value that places a useful share of the resistance on the gas side. The
gas molar flow follows from the ideal-gas density at $50\,\degC$:
$c = p/(RT) = 101\,325/[(8.314)(323)] = 37.7\,\text{mol/m}^3$, so
$G_{\text{mol}} = c\,G = (37.7)(12) = 452\,\text{mol/s}$. On an
inert-gas basis the carrier flow is $G_s = G_{\text{mol}}(1 -
y_{\text{in}}) \approx 451\,\text{mol/s}$.

\subsection{Equilibrium line, operating line, and the solvent rate}

Working in mole ratios $Y = y/(1-y)$ and $X = x/(1-x)$ keeps the carrier
flows constant up the column. The inlet and outlet gas ratios are
$Y_{N+1} = 0.0020/0.9980 = 2.004\times 10^{-3}$ and $Y_1 = 2.0\times
10^{-4}/0.9998 = 2.000\times 10^{-4}$. The solute removed per second is
$G_s(Y_{N+1}-Y_1) = 451(2.004\times 10^{-3} - 2.000\times 10^{-4}) =
451(1.804\times 10^{-3}) = 0.814\,\text{mol/s}$ of \text{SO}$_2$.

The minimum solvent rate is the liquid flow that would leave the bottom
in equilibrium with the entering gas, $X_{\max} = Y_{N+1}/m =
2.004\times 10^{-3}/0.86 = 2.330\times 10^{-3}$. With clean entering
solvent ($X_0 = 0$),
\[
\Big(\frac{L_s}{G_s}\Big)_{\min} = \frac{Y_{N+1}-Y_1}{X_{\max}-X_0} =
\frac{1.804\times 10^{-3}}{2.330\times 10^{-3}} = 0.774.
\]
Operating at $1.4\times$ the minimum, $L_s/G_s = 1.084$, so
$L_s = 1.084(451) = 489\,\text{mol/s}$ of water, about $8.8\,\text{kg/s}$
or $8.8\,\text{L/s}$ of solvent. The actual loaded-solvent ratio is
$X_N = (Y_{N+1}-Y_1)/(L_s/G_s) = 1.804\times 10^{-3}/1.084 =
1.664\times 10^{-3}$, comfortably below the equilibrium $X_{\max}$, which
keeps a finite driving force at the bottom. The absorption factor is
$A = (L_s/G_s)/m = 1.084/0.86 = 1.26$, above unity, so the absorber is
well-behaved and recovery improves with height.
Figure~\ref{fig:vol04ch07:absorption-operating} plots the operating line
sitting above the equilibrium line, with the bottom corner near
equilibrium and the top corner near the clean-gas axis.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-absorption-operating.tex}

\subsection{Number and height of transfer units}

For a dilute absorber with a linear equilibrium line, the number of
overall gas-phase transfer units has the closed form
\[
N_{OG} = \frac{1}{1 - 1/A}\,\ln\!\left[\Big(1 - \frac{1}{A}\Big)
\frac{Y_{N+1}-mX_0}{Y_1-mX_0} + \frac{1}{A}\right].
\]
With $X_0 = 0$, the composition ratio is $(Y_{N+1})/(Y_1) =
2.004\times 10^{-3}/2.000\times 10^{-4} = 10.02$, and $1/A = 0.794$. The
bracket is $(1-0.794)(10.02) + 0.794 = 0.206(10.02) + 0.794 = 2.064 +
0.794 = 2.858$, so
\[
N_{OG} = \frac{1}{0.206}\,\ln(2.858) = 4.854(1.050) = 5.10.
\]
The separation demands about five overall gas-phase transfer units. The
height of a transfer unit packages the overall coefficient and the
packing interfacial area, $H_{OG} = G_s/(K_y a\, S)$, and for a
well-irrigated random packing in this service a working value is
$H_{OG} = 0.75\,\text{m}$. The packed height is then
\[
Z = H_{OG}\,N_{OG} = (0.75)(5.10) = 3.8\,\text{m}.
\]
We carry $4.0\,\text{m}$ of packing to leave margin against the
uncertainty in $H_{OG}$, which is the least certain number in the chain
and would be confirmed against a packing-vendor correlation or a pilot
run before construction.

\subsection{Column diameter from flooding}

The diameter is set by the gas throughput against the flooding limit of
the packing. The flow parameter is
\[
F_{LV} = \frac{L}{G}\sqrt{\frac{\rho_G}{\rho_L}},
\]
where $L$ and $G$ here are mass flows. The gas mass flow is
$G_{\text{mol}}$ times the mean molar mass of flue gas
($\approx 0.030\,\text{kg/mol}$), $\dot m_G = (452)(0.030) =
13.6\,\text{kg/s}$; the liquid mass flow is $\dot m_L = 8.8\,\text{kg/s}$.
With $\rho_G = c\,M = (37.7)(0.030) = 1.13\,\text{kg/m}^3$ and
$\rho_L = 1000\,\text{kg/m}^3$,
\[
F_{LV} = \frac{8.8}{13.6}\sqrt{\frac{1.13}{1000}} = (0.647)(0.0336) =
0.0217.
\]
On the generalised pressure-drop correlation of
figure~\ref{fig:vol04ch07:flooding}, a flow parameter near $0.02$ and a
target pressure drop of $\sim 300\,\text{Pa/m}$ place the capacity
parameter at roughly $C_S = 0.07\,\text{m/s}$ (the corrected superficial
gas velocity grouping).

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-flooding.tex}

Translating to a flooding-fraction design, a
working superficial gas velocity for $50$-mm metal random packing at
$60$ to $70\,\%$ of flood in this service is $u_g \approx 2.2\,\text{m/s}$.
The column cross-section is
\[
S = \frac{G}{u_g} = \frac{12}{2.2} = 5.45\,\text{m}^2,
\quad D = \sqrt{\frac{4S}{\pi}} = \sqrt{\frac{4(5.45)}{\pi}} =
2.6\,\text{m}.
\]
A $2.6\,\text{m}$ shell carries the gas at the chosen flooding fraction;
rounding up to a standard $2.7\,\text{m}$ vessel drops the velocity to
$u_g = 12/5.73 = 2.1\,\text{m/s}$ and buys a little more margin against
flood.

\subsection{Pressure drop and the verdict}

At the design point on the $300\,\text{Pa/m}$ contour, the irrigated
pressure drop across $4.0\,\text{m}$ of packing is
$\Delta p \approx (300)(4.0) = 1200\,\text{Pa}$, about
$0.012\,\text{atm}$, plus a few hundred pascal across the distributor
and demister. A draft fan must supply on the order of $1.5\,\text{kPa}$
to push the $12\,\text{m}^3/\text{s}$ of gas through the column; the fan
power is $\dot W \approx \dot V\,\Delta p/\eta = (12)(1500)/0.7 =
2.6\times 10^{4}\,\text{W} \approx 26\,\text{kW}$, a modest parasitic
load on the boiler. The verdict: a packed column roughly $2.7\,\text{m}$
in diameter with $4.0\,\text{m}$ of random packing, irrigated at
$8.8\,\text{L/s}$ of dilute caustic, removes $90\,\%$ of the \text{SO}$_2$
and meets the $200\,\text{ppm}$ stack limit, at a fan penalty near
$26\,\text{kW}$ and a packed-section pressure drop near $1.2\,\text{kPa}$.

The dominant uncertainties, in order, are the equilibrium slope $m$ (set
by the caustic chemistry and the operating loading, and the dominant
control on the solvent rate and on $N_{OG}$), the height of a transfer
unit $H_{OG}$ (set by the packing and the irrigation rate, and the
direct multiplier on the packed height), and the flooding velocity (set
by the packing geometry and the physical properties, and the dominant
control on the diameter). A
real design would pin $m$ from a vapour-liquid equilibrium correlation
for the caustic system at the operating loading, pin $H_{OG}$ from a
packing-vendor mass-transfer correlation or a pilot column, and pin the
flooding velocity from the same vendor's capacity chart, then re-run the
chain. The arithmetic here is the first cut that tells the engineer the
column is a few metres across and a few metres tall, not a few
centimetres or a few tens of metres, which is the judgment the
back-of-envelope is for \cite{text:treybal-mass-transfer}.

\section{A second case study: a seawater reverse-osmosis unit}\pathtag{standard}

The scrubber sized a gas-liquid contactor against a transfer-unit
analysis; a reverse-osmosis train sizes a pressure-driven membrane
against an osmotic back-pressure and an energy budget, and the two
together span the range of separation design. We take a small
seawater-desalination skid and carry it from the feed salinity through
the osmotic pressure, the net driving pressure, the water flux, the
membrane area, the recovery limit, and the specific energy with and
without an energy-recovery device, ending with a verdict an engineer
would sign. The membrane appears in this volume once as a transport
barrier; here it appears as a costed unit operation, and the controlling
numbers are the ones a desalination engineer carries.

\subsection{The feed and the duty}

A coastal plant must produce $P = 1200\,\text{m}^3/\text{day}$ of potable
permeate from seawater of total dissolved solids $S_f = 35\,\text{g/L}$
at $25\,\degC$. The permeate target is $S_p \le 0.5\,\text{g/L}$, well
inside drinking-water guidance, which a good seawater membrane meets at a
salt rejection above $99.5\,\%$. Converting the production rate to working
units, $P = 1200/86\,400 = 1.39\times 10^{-2}\,\text{m}^3/\text{s} =
50\,\text{m}^3/\text{h}$. The recovery, the permeate fraction of the
feed, is chosen at $r = 0.45$, the value near which the specific energy of
a single seawater pass is least, so the feed flow is $F = P/r =
1200/0.45 = 2670\,\text{m}^3/\text{day}$ and the reject brine carries
the balance, $B = F - P = 1470\,\text{m}^3/\text{day}$.

\subsection{Osmotic pressure and the net driving pressure}

The osmotic pressure of seawater follows from the van~'t~Hoff relation
$\pi = i\,c\,RT$, where $c$ is the molar salt concentration and $i$ the
dissociation count. For sodium chloride as a stand-in, $35\,\text{g/L}$
at a molar mass of $58.4\,\text{g/mol}$ gives $c = 0.599\,\text{mol/L} =
599\,\text{mol/m}^3$, and with $i = 2$ (sodium and chloride),
\[
\pi_f = i c R T = (2)(599)(8.314)(298) = 2.97\times 10^{6}\,\text{Pa}
\approx 2.7\,\text{MPa},
\]
matching the working seawater figure once the slight non-ideality of real
seawater is folded in. The osmotic pressure is not the feed value across
the whole module: as water permeates, the brine concentrates, so by the
reject end the salt is more concentrated by $1/(1-r)$ and the local
osmotic pressure has climbed. At $r = 0.45$ the reject osmotic pressure is
\[
\pi_b = \frac{\pi_f}{1-r} = \frac{2.7}{0.55} = 4.9\,\text{MPa},
\]
and the design must beat this larger back-pressure, not the feed value,
to keep producing permeate at the reject end. Taking a feed pressure
$\Delta p = 7.0\,\text{MPa}$, the net driving pressure at the feed end is
$7.0 - 2.7 = 4.3\,\text{MPa}$ and at the reject end is $7.0 - 4.9 =
2.1\,\text{MPa}$; the log-mean net driving pressure across the module is
\[
\overline{\Delta p_{\text{net}}} = \frac{4.3 - 2.1}{\ln(4.3/2.1)} =
\frac{2.2}{\ln 2.05} = \frac{2.2}{0.716} = 3.1\,\text{MPa}.
\]
The brine-end driving pressure of $2.1\,\text{MPa}$ stays positive, so
the module produces permeate along its whole length;
figure~\ref{fig:vol04ch07:ro-recovery} plots the osmotic climb against
recovery and shows how the brine-end driving force is what runs out
first as recovery is pushed up.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-ro-recovery.tex}

\subsection{Flux, membrane area, and the module count}

The water flux through the membrane is the permeability times the net
driving pressure, $J_w = A_w(\overline{\Delta p_{\text{net}}})$, with a
working seawater-membrane permeability $A_w = 1.0\times 10^{-12}\,\text{m}
/(\text{s}\cdot\text{Pa})$. At the log-mean driving pressure,
\[
J_w = (1.0\times 10^{-12})(3.1\times 10^{6}) = 3.1\times 10^{-6}\,
\text{m/s} = 11\,\text{L}/(\text{m}^2\cdot\text{h}),
\]
in the expected seawater range, lower than a brackish unit because the
osmotic back-pressure eats most of the applied pressure. The membrane
area to carry the permeate is the volumetric permeate rate divided by the
flux,
\[
A_m = \frac{P}{J_w} = \frac{1.39\times 10^{-2}}{3.1\times 10^{-6}} =
4.5\times 10^{3}\,\text{m}^2.
\]
At $37\,\text{m}^2$ per standard $8\,\text{inch}$ spiral-wound element,
the train needs about $A_m/37 = 121$ elements, packaged seven to a
pressure vessel, so roughly $17$ to $18$ vessels arranged in a two-stage
taper that keeps the brine velocity up as the flow falls. The
concentration-polarisation factor, the ratio of wall to bulk salt
concentration, sits near $1.1$ to $1.2$ in a well-designed cross-flow
element; folding it in raises the effective wall osmotic pressure and
trims the flux by roughly that fraction, which is why the area carries a
margin rather than sitting at the bare-flux value.

\subsection{Specific energy and the recovery device}

The pump work sets the operating cost. The ideal pump power to pressurise
the feed is $\dot W = F\,\Delta p$, with $F$ the volumetric feed rate. In
consistent units $F = 2670/86\,400 = 3.09\times 10^{-2}\,\text{m}^3/
\text{s}$, so
\[
\dot W_{\text{feed}} = F\,\Delta p = (3.09\times 10^{-2})(7.0\times
10^{6}) = 2.16\times 10^{5}\,\text{W} = 216\,\text{kW}.
\]
Without energy recovery, the specific energy per cubic metre of permeate
is $\dot W_{\text{feed}}/P = 2.16\times 10^{5}/(1.39\times 10^{-2}) =
1.56\times 10^{7}\,\text{J/m}^3 = 4.3\,\text{kWh/m}^3$ at a pump
efficiency of unity, and near $6\,\text{kWh/m}^3$ once a $0.7$ to $0.75$
pump-and-motor efficiency is folded in. The reject brine leaves the module
at nearly the feed pressure, carrying $B = 1470\,\text{m}^3/\text{day}$ at
$\sim 6.8\,\text{MPa}$ of recoverable pressure. An energy-recovery device
(an isobaric pressure exchanger or a turbine) returns most of that brine
pressure to the incoming feed. The brine fraction is $1 - r = 0.55$, so a
device at $95\,\%$ transfer efficiency recovers about
$0.55\times 0.95 = 0.52$ of the feed pressurisation energy, cutting the
specific energy to roughly $4.3(1 - 0.52) \approx 2.1\,\text{kWh/m}^3$ at
the ideal pump and near $3\,\text{kWh/m}^3$ with real efficiencies, the
modern seawater figure cited in section~7.6. The energy-recovery device
is the single component that moves seawater desalination from the
$6\,\text{kWh/m}^3$ of the 1990s to the $3\,\text{kWh/m}^3$ of today,
and figure~\ref{fig:vol04ch07:ro-recovery} marks why the recovery is held
near $0.45$: pushing it higher drives the brine osmotic pressure toward
the applied pressure and the specific energy back up.

\subsection{The verdict}

A seawater train of about $4500\,\text{m}^2$ of membrane, $120$-odd
spiral-wound elements in $17$ to $18$ pressure vessels on a two-stage
taper, fed at $7\,\text{MPa}$ and $45\,\%$ recovery with an isobaric
energy-recovery device, delivers $1200\,\text{m}^3/\text{day}$ of
$0.5\,\text{g/L}$ permeate at a specific energy near $3\,\text{kWh/m}^3$.
The dominant uncertainties, in order, are the membrane permeability
$A_w$ (set by the chosen element and ageing in service, the direct
multiplier on area), the concentration-polarisation factor (set by the
cross-flow hydraulics, which raises the wall osmotic pressure above the
bulk and trims the flux), and the pump-and-recovery-device efficiency
chain (the direct control on the energy figure). A real design would pin
$A_w$ from the element datasheet at end-of-life rather than new, pin the
polarisation factor from the element's feed-spacer hydraulics, and pin the
efficiencies from the pump and pressure-exchanger curves, then re-run the
chain. The arithmetic here is the first cut that tells the engineer the
plant is a few thousand square metres of membrane and a few hundred
kilowatts of pump, not a few hundred square metres or a few megawatts,
which is the judgment the back-of-envelope is for
\cite{gen:mackay2009,text:perry-chemical-engineers}.

\section{A third case study: a pressure-swing-adsorption oxygen unit}\pathtag{standard}

The scrubber sized a gas-liquid contactor against a transfer-unit
analysis and the reverse-osmosis train sized a pressure-driven membrane
against an osmotic budget; a pressure-swing-adsorption unit sizes a
gas-solid contactor against an isotherm and a cycle, and the three together
span the unit-operation range of the chapter. We take a small medical-grade
oxygen concentrator that pulls nitrogen out of air on a zeolite molecular
sieve, and carry it from the feed and product specifications through the
isotherm working capacity, the sorbent inventory, the bed dimensions from
the cycle time, the blower power, and a verdict an engineer would sign.
Adsorption appears in section~7.5 as a unit operation governed by an
isotherm and a breakthrough curve; here it appears as a costed cyclic
process, and the controlling numbers are the ones an oxygen-plant engineer
carries.

\subsection{The feed, the product, and the sorbent}

A bedside concentrator must deliver $P = 5\,\text{L/min}$ of
$90\,\%$-oxygen product gas from ambient air ($21\,\%$ oxygen,
$78\,\%$ nitrogen, $1\,\%$ argon) at sea level. Zeolite 13X adsorbs
nitrogen far more strongly than oxygen, so passing compressed air over the
sieve holds nitrogen on the solid and lets an oxygen-enriched stream pass;
argon, which the sieve treats like oxygen, sets the purity ceiling near
$95\,\%$ for a single sieve, comfortably above the $90\,\%$ target. The
unit runs the four-step Skarstrom cycle of
figure~\ref{fig:vol04ch07:psa-cycle}: pressurise the bed to the high
pressure $p_H$, pass feed and withdraw oxygen product at $p_H$, blow the
bed down to the low pressure $p_L$ to release the held nitrogen, and purge
with a slip of product oxygen to sweep the desorbed nitrogen out before the
next pressurisation. Two beds run a half-period out of phase so the product
flow is continuous. Take $p_H = 5\,\text{bar}$ absolute and $p_L =
1\,\text{bar}$ absolute, a vacuum-free swing suited to a small machine.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-psa-cycle.tex}

\subsection{Working capacity from the isotherm}

The separation is paid for by the nitrogen the sieve loads at $p_H$ and
releases at $p_L$, the working capacity. The nitrogen isotherm on 13X is
favourable and well fitted by a Langmuir form
$q = q_m b\,p_{\text{N}_2}/(1 + b\,p_{\text{N}_2})$ with a saturation
loading $q_m = 3.0\,\text{mol/kg}$ and an affinity $b = 0.9\,\text{bar}^{-1}$
on the nitrogen partial pressure. At $p_H = 5\,\text{bar}$ the feed
nitrogen partial pressure is $0.78(5) = 3.9\,\text{bar}$, giving a loading
\[
q_H = \frac{(3.0)(0.9)(3.9)}{1 + (0.9)(3.9)} = \frac{10.53}{4.51} =
2.34\,\text{mol/kg};
\]
at $p_L = 1\,\text{bar}$, during the purge the bed sees nitrogen partial
pressure near the desorbing value, and taking a representative
$p_{\text{N}_2}\approx 0.5\,\text{bar}$ at the low-pressure end gives
\[
q_L = \frac{(3.0)(0.9)(0.5)}{1 + (0.9)(0.5)} = \frac{1.35}{1.45} =
0.93\,\text{mol/kg}.
\]
The working capacity is the swing $\Delta q = q_H - q_L = 2.34 - 0.93 =
1.41\,\text{mol/kg}$ of nitrogen removed per kilogram of sieve per cycle,
the quantity figure~\ref{fig:vol04ch07:psa-isotherm} reads off the isotherm
as the vertical interval between the two operating pressures. The favourable
isotherm helps and hurts at once: it loads heavily at modest pressure, but
it flattens at high pressure, so most of the swing comes from the lower
part of the curve and pushing $p_H$ past $5\,\text{bar}$ buys little extra
capacity while costing compressor work.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-psa-isotherm.tex}

\subsection{Sorbent inventory and the bed size}

The nitrogen throughput sets the sorbent mass. To make $5\,\text{L/min}$ of
$90\,\%$ oxygen the unit must reject the nitrogen that accompanies the
oxygen in the feed. A nitrogen balance on the product specification: each
mole of product carries $0.90$ oxygen and $0.10$ nitrogen-plus-argon, and
the feed that supplies $0.90$ mol of oxygen carries $0.90(0.78/0.21) =
3.34$ mol of nitrogen, of which all but the $0.10$ slip must be adsorbed,
about $3.24$ mol of nitrogen captured per mole of product oxygen. The
product molar rate is $5\,\text{L/min}$ at the product condition, about
$5/22.4 = 0.223\,\text{mol/min}$ at standard conditions, so $0.201\,
\text{mol/min}$ of oxygen and roughly $0.201(3.24)/0.90 = 0.72\,\text{mol/
min}$ of nitrogen captured, $1.2\times 10^{-2}\,\text{mol/s}$. Over a
half-cycle adsorption step of $t_{\text{ads}} = 10\,\text{s}$, each bed
must hold the nitrogen captured in that window, $n_{\text{N}_2} =
(1.2\times 10^{-2})(10) = 0.12\,\text{mol}$ per cycle per bed. At the
working capacity $\Delta q = 1.41\,\text{mol/kg}$ the sorbent mass per bed
is
\[
m_{\text{sieve}} = \frac{n_{\text{N}_2}}{\Delta q} = \frac{0.12}{1.41} =
0.085\,\text{kg},
\]
call it $0.1\,\text{kg}$ per bed with margin against incomplete utilisation
of the bed. At a packed-bed bulk density of $\rho_b = 700\,\text{kg/m}^3$
for 13X beads, the bed volume is $V_b = 0.1/700 = 1.4\times 10^{-4}\,
\text{m}^3 = 140\,\text{cm}^3$, a cylinder roughly $4\,\text{cm}$ in
diameter and $11\,\text{cm}$ tall, the size of a soda can, which matches the
sieve canisters of real bedside concentrators. The bed length must also be
long enough that the nitrogen mass-transfer zone fits inside it before
breakthrough; for fast-cycling 13X the zone is a few centimetres, so the
$11\,\text{cm}$ bed carries the zone with room to spare, and the limit here
is capacity, not zone length.

\subsection{Blower power and the verdict}

The compressor work sets the operating cost and the noise budget. The unit
pressurises the feed air from $1\,\text{bar}$ to $p_H = 5\,\text{bar}$. The
feed air rate is the product oxygen rate scaled up by the recovery: a
single-stage PSA recovers roughly half the oxygen in the feed. On a
per-second basis the product oxygen rate is $0.201\,\text{mol/min} =
3.35\times 10^{-3}\,\text{mol/s}$, so the air intake is
$3.35\times 10^{-3}/(0.21\times 0.5) = 3.2\times 10^{-2}\,\text{mol/s}$
of air, about $0.72\,\text{L/s}$ at the intake. The ideal
isothermal compression work for an ideal gas is $\dot W = \dot n RT\,
\ln(p_H/p_L)$,
\[
\dot W = (3.2\times 10^{-2})(8.314)(298)\,\ln 5 = (3.2\times 10^{-2})
(2478)(1.609) = 128\,\text{W},
\]
and at a small-compressor efficiency near $0.5$ the shaft draw is about
$260\,\text{W}$, with the rest of the wall-plug load going to the valves,
controls, and cooling, landing the machine near $300\,\text{to }350\,
\text{W}$, the rated draw of a bedside concentrator. The verdict: two
sieve beds of about $0.1\,\text{kg}$ of 13X each, roughly $4\,\text{cm}$ by
$11\,\text{cm}$, cycled on a $20\,\text{s}$ Skarstrom cycle between
$1\,\text{and }5\,\text{bar}$, deliver $5\,\text{L/min}$ of $90\,\%$ oxygen
at a wall-plug draw near $300\,\text{W}$.

The dominant uncertainties, in order, are the nitrogen working capacity
$\Delta q$ (set by the isotherm at the operating temperature and the actual
low-pressure nitrogen partial pressure reached in the purge, and the direct
divisor on the sorbent mass), the oxygen recovery (set by how much product
gas is spent on the purge and pressurisation, and the direct multiplier on
the blower air rate and power), and the cycle time (set by the
mass-transfer-zone speed and the valve switching, and the control on how
much sorbent the unit needs). A real design would pin $\Delta q$ from a
measured isotherm at the bed operating temperature including the heat of
adsorption that warms the bed and lowers capacity, pin the recovery from a
cyclic simulation of the four steps, and pin the cycle time from a
breakthrough run, then re-run the chain. The arithmetic here is the first
cut that tells the engineer the machine is a soda-can of sieve and a few
hundred watts of blower, not a drum of sieve or a few kilowatts, which is
the judgment the back-of-envelope is for
\cite{text:treybal-mass-transfer,text:perry-chemical-engineers}.

\section{An applied vignette: a membrane blood oxygenator}\pathtag{mastery}

The three case studies sized industrial separators; the same two-film
algebra sizes a device that stands in for a human lung during open-heart
surgery, and the change of context sharpens the lesson that the slowest
film sets the rate. A membrane blood oxygenator passes blood on one side
of a bundle of microporous hollow fibres and oxygen on the other,
transferring oxygen into the blood and carbon dioxide out, with the
membrane keeping the gas and the blood from mixing directly. We size the
fibre area for a patient-scale unit and find where the resistance sits,
which is the design question that decides the device.

\subsection*{The duty and the driving force}

An adult on full cardiopulmonary bypass needs an oxygen transfer of about
$\dot n_{\text{O}_2} = 250\,\text{mL/min}$ at standard conditions, which is
$250/(22\,400) = 1.12\times 10^{-2}\,\text{mol/min} = 1.86\times 10^{-4}\,
\text{mol/s}$. Blood enters venous, with an oxygen partial pressure near
$40\,\text{mmHg}$, and must leave arterial, near $100\,\text{mmHg}$; the
gas side runs pure oxygen at close to $760\,\text{mmHg}$. The driving force
is the partial-pressure gap between the gas and the blood, and because the
blood oxygen rises along the device, the log-mean of the inlet and outlet
gaps is the right average. At the inlet the gap is $760-40 = 720\,\text{
mmHg}$ and at the outlet $760-100 = 660\,\text{mmHg}$, so the log-mean is
$(720-660)/\ln(720/660) = 60/0.0870 = 690\,\text{mmHg} = 9.2\times 10^{4}
\,\text{Pa}$. Oxygen solubility in blood folds the partial pressure into a
concentration through an effective Henry coefficient; for the sizing we
carry the transfer as a coefficient times a partial-pressure difference,
$\dot n_{\text{O}_2} = K_o\,A\,\Delta p_{\text{lm}}$, with $K_o$ a
permeance in $\text{mol}/(\text{m}^2\cdot\text{s}\cdot\text{Pa})$.

\subsection*{The resistance in series and the fibre area}

The oxygen crosses three resistances: the blood-side film, the membrane
wall, and the gas-side film. For a microporous fibre running pure oxygen,
the gas-side and membrane resistances are small against the blood film,
because oxygen diffuses fast in the gas and through the open pores but
slowly through plasma and against the slow uptake by haemoglobin inside
the red cells. A representative overall permeance for a commercial fibre
bundle is $K_o = 3.0\times 10^{-9}\,\text{mol}/(\text{m}^2\cdot\text{s}
\cdot\text{Pa})$, blood-side controlled. The fibre area follows from the
duty,
\[
A = \frac{\dot n_{\text{O}_2}}{K_o\,\Delta p_{\text{lm}}} =
\frac{1.86\times 10^{-4}}{(3.0\times 10^{-9})(9.2\times 10^{4})} =
0.67\,\text{m}^2.
\]
Commercial adult oxygenators carry $1.5$ to $2.5\,\text{m}^2$ of fibre,
above this bare estimate, and the gap is the design margin: the blood film
is the controlling resistance, and it worsens wherever the blood channels
or stagnates between fibres, so the device is built with several times the
ideal area to hold its transfer when the flow is less than perfectly
distributed and as a plasma layer builds on the fibres in a long run.
Figure~\ref{fig:vol04ch07:oxygenator} draws the fibre bundle and the
oxygen profile across the blood film, with the steep drop on the blood
side that marks where the resistance lives.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-oxygenator.tex}

\subsection*{The carbon-dioxide check and the verdict}

Carbon dioxide leaves the blood about twenty times more readily than
oxygen enters it, because it is far more soluble and diffuses through
plasma with a larger effective coefficient, so the area sized for oxygen
clears carbon dioxide with room to spare; the sweep-gas flow, not the
fibre area, sets the carbon-dioxide removal, and the operator trims it to
hold the arterial carbon-dioxide partial pressure. The verdict: a fibre
bundle near $2\,\text{m}^2$, blood on the shell side at a few litres per
minute and pure oxygen through the lumens, transfers the
$250\,\text{mL/min}$ of oxygen a bypassed patient needs, with the blood
film as the controlling resistance and the area carrying a safety factor
against maldistribution and plasma fouling. The same Krogh-style balance
that set the tissue oxygen reach in the worked example runs here in
reverse, with the fibre playing the role of the capillary and the blood
the role of the tissue; the device is a lung built from the two-film model
\cite{text:bird-transport-phenomena}.

\section{A second applied vignette: electrodialysis of brackish
water}\pathtag{mastery}

The pressure-driven membrane of the reverse-osmosis case study moves water
against the salt; an electrodialysis stack moves the salt against the water,
driven by an electric field rather than a pressure, and the change of driving
force turns the economics on their head for a lightly salted feed. Where
reverse osmosis pays to pressurise the whole stream and pushes the abundant
water through the membrane, electrodialysis pays only to drag the scarce ions
across, so it wins wherever the salt is a small fraction of the feed. We size
a small electrodialysis unit to bring a brackish groundwater down to
drinking-water salinity and find where its limit sits.

\subsection*{The stack and the driving force}

A brackish feed at $S_f = 3.0\,\text{g/L}$ total dissolved solids must be
desalted to $S_p = 0.5\,\text{g/L}$. The stack is built of alternating
cation-exchange and anion-exchange membranes between one electrode pair, as
figure~\ref{fig:vol04ch07:electrodialysis} draws it: under the applied
potential cations migrate toward the cathode and pass the cation-exchange
membranes but are stopped by the anion-exchange ones, and anions do the
mirror image, so ions are pumped out of every other channel (the dilute
channels) and trapped in the channels between (the concentrate channels). A
cell pair is one membrane of each kind and their two channels, and a
production stack carries several hundred cell pairs so that a single electrode
current desalts every dilute channel at once, which is what makes the
electrode losses, incurred only once per stack, a small overhead.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-electrodialysis.tex}

The charge that must be moved is set by the salt to be removed, through
Faraday's law. Removing salt from $3.0$ to $0.5\,\text{g/L}$ takes out
$2.5\,\text{g/L}$, and at a mean equivalent weight near $58\,\text{g/eq}$ for
the mixed salt that is $2.5/58 = 4.3\times 10^{-2}\,\text{eq/L}$ of ions to
transport. For a target product rate of $Q_p = 2.0\,\text{m}^3/\text{h} =
5.6\times 10^{-4}\,\text{m}^3/\text{s} = 0.56\,\text{L/s}$, the ionic
transport rate is $(4.3\times 10^{-2})(0.56) = 2.4\times 10^{-2}\,\text{eq/s}$.
Each equivalent carries one Faraday, $F = 96\,485\,\text{C/eq}$, so the ideal
stack current, if every electron moved exactly one ion across one cell pair,
divides the charge rate by the number of cell pairs $N_{\text{cp}}$. Taking
$N_{\text{cp}} = 200$ cell pairs and a current efficiency $\eta_i = 0.90$
(some current is lost to co-ion leakage and water splitting),
\[
I = \frac{F\,(\text{eq/s})}{N_{\text{cp}}\,\eta_i} =
\frac{(96\,485)(2.4\times 10^{-2})}{(200)(0.90)} =
\frac{2316}{180} = 12.9\,\text{A}.
\]
A stack current near $13\,\text{A}$ desalts the whole product flow, a modest
current because the ions are dragged one cell pair at a time across two
hundred pairs in electrical series.

\subsection*{The limiting current and the verdict}

Electrodialysis carries a hard ceiling that reverse osmosis does not: the
limiting current density, the current at which the dilute-channel salt at the
membrane face runs to zero and no more ions can be delivered to carry the
charge. Beyond it the current is forced to split water into hydrogen and
hydroxide ions, wasting energy and swinging the local pH enough to scale the
membranes. The limiting current density is set by the mass transfer of salt
to the membrane face,
\[
i_{\text{lim}} = \frac{F\,k\,c_d}{t_{\text{cm}} - t_{\text{sol}}},
\]
with $k$ the dilute-channel mass-transfer coefficient, $c_d$ the local
dilute concentration, and $t_{\text{cm}} - t_{\text{sol}}$ the difference
between the counter-ion transport number in the membrane (near unity) and in
solution (near one-half). The ceiling falls as the dilute channel is
desalted, because $c_d$ drops, so the stack must run at declining current
density down its length and the last, most-desalted channels set the limit.
For a working $k = 2\times 10^{-5}\,\text{m/s}$, a dilute exit concentration
$c_d = 0.5/58 = 8.6\times 10^{-3}\,\text{eq/L} = 8.6\,\text{eq/m}^3$, and
$t_{\text{cm}} - t_{\text{sol}} = 0.4$,
\[
i_{\text{lim}} = \frac{(96\,485)(2\times 10^{-5})(8.6)}{0.4} =
\frac{16.6}{0.4} = 41\,\text{A/m}^2,
\]
so the exit end tolerates about $40\,\text{A/m}^2$; the design runs safely
below it, near $30\,\text{A/m}^2$, to leave margin against maldistribution.
At that current density the membrane area per cell pair to pass the
$12.9\,\text{A}$ is $I/(N_{\text{cp}}\,i) = 12.9/(200 \times 30) = 2.2\times
10^{-3}\,\text{m}^2$ per pair, so a cell pair of about $22\,\text{cm}$ by
$10\,\text{cm}$ carries it. The verdict: a stack of two hundred cell pairs of
palm-sized membranes, drawing about $13\,\text{A}$ at the stack voltage,
desalts brackish water at $2\,\text{m}^3/\text{h}$, and the design constraint
that has no analogue in the pressure-driven case is the limiting current at
the desalted exit, which caps how hard the last channels can be pushed and is
the reason electrodialysis is chosen for brackish rather than seawater feeds:
the energy scales with the salt removed, so it is cheap at low salinity and
uncompetitive with reverse osmosis once the feed is as salty as the sea.

\section{A third applied vignette: a transdermal drug-delivery
patch}\pathtag{mastery}

The separators of this chapter moved species out of a stream; a transdermal
patch runs the same diffusion physics to move a drug at a controlled rate into
a patient across the skin, and the design question is the mirror of a
separator's: not how fast can the flux be pushed, but how steadily can it be
held at a therapeutic value for days. The patch is a diffusion problem with the
skin as the membrane, and the two-resistance and steady-flux relations of the
chapter size it directly.

\subsection*{The barrier and the target dose}

The rate-limiting barrier is the stratum corneum, the dead outermost skin layer
some $L = 20\,\si{\micro\meter}$ thick, through which the drug must diffuse
before it reaches the capillaries below and is swept into the blood. A steady
dose calls for a steady flux, and a reservoir patch is built to hold the drug
concentration at its upstream face constant so that the flux settles at the
Fick steady value across the skin,
\[
J = \frac{D_s\,K\,C_v}{L},
\]
with $D_s$ the drug diffusivity in the stratum corneum, $K$ the
skin-to-vehicle partition coefficient, and $C_v$ the drug concentration in the
patch reservoir. The grouping $P = D_s K/L$ is the skin permeability, the
number a formulator measures once and carries. Take a lipophilic drug with a
measured skin permeability $P = 2.0\times 10^{-7}\,\text{m/s}$ against water
as the reference, a reservoir concentration $C_v = 5.0\,\text{kg/m}^3$, and a
target delivery of $5\,\text{mg}$ over $24\,\text{h}$, which is
$5\times 10^{-6}/86\,400 = 5.8\times 10^{-11}\,\text{kg/s}$ of drug. The flux
the skin passes is $J = P\,C_v = (2.0\times 10^{-7})(5.0) = 1.0\times 10^{-6}\,
\text{kg}/(\text{m}^2\cdot\text{s})$, so the patch area is
\[
A = \frac{\dot m}{J} = \frac{5.8\times 10^{-11}}{1.0\times 10^{-6}} =
5.8\times 10^{-5}\,\text{m}^2 = 0.58\,\text{cm}^2,
\]
a patch under a square centimetre, which matches the size of a
low-dose nicotine or hormone patch. A drug of low skin permeability would need
a proportionally larger patch, and a permeability below about $10^{-8}\,
\text{m/s}$ pushes the patch area past what the skin can comfortably carry,
which is why transdermal delivery is limited to potent, lipophilic,
small-molecule drugs and why permeation enhancers, which raise $D_s$ or $K$,
are formulated in to lift a marginal candidate over the area ceiling.

\subsection*{The lag, the burst, and the verdict}

The steady flux is not reached instantly. The stratum corneum must load its
diffusion profile first, and the delay is the same time lag the polymer-film
permeation example computed, $t_{\text{lag}} = L^2/(6 D_s)$. With a stratum-
corneum diffusivity $D_s \approx 1\times 10^{-13}\,\si{\meter\squared\per
\second}$ and $L = 20\,\si{\micro\meter}$, the lag is $t_{\text{lag}} =
(2.0\times 10^{-5})^2/[6(1\times 10^{-13})] = 4.0\times 10^{-10}/6.0\times
10^{-13} = 667\,\text{s} \approx 11\,\text{min}$ if the reservoir concentration
is stepped on, but the skin's own tortuosity stretches the practical lag to an
hour or more, so a patch takes time to reach its therapeutic flux and must be
worn ahead of need. The mirror problem is a matrix patch that has been stored
long enough for the drug to diffuse into the adhesive and saturate the skin
contact, which delivers an initial burst above the steady rate. The verdict: a
sub-square-centimetre reservoir patch delivers a milligram-scale daily dose at
a flux the skin permeability and the reservoir concentration set, with a lag of
order an hour before the steady flux is reached and a design constraint that
the skin, not the patch, is the rate-limiting barrier. The same steady-flux and
time-lag relations that governed the food-packaging film and the membrane
sensor govern the patch; the patch is a diffusion cell with the patient's
capillaries as the downstream sink \cite{text:bird-transport-phenomena}.

\section{A fourth applied vignette: pervaporation past the ethanol-water
azeotrope}\pathtag{mastery}

The chapter's estimation block ends on the wall that a binary distillation
cannot cross the ethanol-water azeotrope at $89\,\text{mol-}\%$ ethanol, and
that dehydrated ethanol needs a second unit operation to get past it.
Pervaporation is that operation done as a membrane process, and it sizes
against a partial-pressure driving force the same way the gas-separation
membrane did, with a phase change built into the transport. The vignette closes
the loop the estimation block opened.

\subsection*{Why the azeotrope is no barrier to a membrane}

Distillation separates by the vapour-liquid equilibrium, and at the azeotrope
that equilibrium gives a vapour of the same composition as the liquid, so no
further enrichment is possible by boiling. A membrane separates by the ratio of
permeabilities through the membrane material, which has nothing to do with the
vapour-liquid equilibrium, so a water-selective membrane pulls water
preferentially out of an azeotropic mixture even though boiling cannot.
Pervaporation feeds the liquid mixture to one face of a dense hydrophilic
membrane, holds the permeate side under vacuum, and lets the permeating species
evaporate as it crosses, so the driving force is the difference between the
species' partial pressure in equilibrium with the feed liquid and its low
partial pressure on the evacuated permeate side. The water, far more permeable
through the hydrophilic membrane than the ethanol, crosses preferentially and
is condensed on the permeate side, and the retentate leaves drier than the
azeotrope.

\subsection*{Sizing the dehydration duty}

We dehydrate an azeotropic feed at $89\,\text{mol-}\%$ ethanol, $11\,\text{mol-
}\%$ water, down to $99.5\,\text{mol-}\%$ ethanol product, a common fuel-grade
target, at a feed rate of $F = 1.0\,\text{mol/s}$. The water to remove is the
difference between the feed and product water flows; taking the ethanol as
nearly conserved in the retentate, the product carries $0.89\,\text{mol/s}$ of
ethanol at $99.5\,\%$, so its water is $0.89(0.005/0.995) = 4.5\times 10^{-3}\,
\text{mol/s}$, and the water permeated is $0.11 - 4.5\times 10^{-3} =
0.106\,\text{mol/s}$, about $1.9\,\text{g/s}$. The membrane area follows from a
water flux; a working hydrophilic pervaporation membrane passes water at
$J_w \approx 1.0\,\text{kg}/(\text{m}^2\cdot\text{h}) = 2.8\times 10^{-4}\,
\text{kg}/(\text{m}^2\cdot\text{s})$ at this water activity, so the area is
\[
A = \frac{\dot m_w}{J_w} = \frac{1.9\times 10^{-3}}{2.8\times 10^{-4}} =
6.8\,\text{m}^2
\]
per mole per second of feed, a modest membrane inventory. The energy cost is
the latent heat the permeating water carries away, drawn from the feed, so the
feed cools as it is dehydrated and must be reheated between stages; the latent
duty is $\dot m_w h_{fg} = (1.9\times 10^{-3})(2.26\times 10^{6}) =
4.3\times 10^{3}\,\text{W}$, about $4\,\text{kW}$ per mole per second of feed,
far below the reboiler duty a distillation would spend to chase the same last
few percent of water toward the azeotrope. The verdict: a few square metres of
hydrophilic membrane per mole per second of feed, under vacuum on the permeate
side with interstage reheat to replace the latent heat, dehydrates ethanol
past the azeotrope at a fraction of the energy an azeotropic or extractive
distillation would spend. Pervaporation is chosen exactly where the estimation
block said distillation runs into the wall: the last approach to a pure product
across an azeotrope, where the membrane's permeability ratio does the work the
vanishing relative volatility cannot \cite{text:perry-chemical-engineers}.

\section{A fifth applied vignette: a simulated moving bed for a chiral
separation}\pathtag{mastery}

The adsorption case study ran a batch bed to breakthrough and then regenerated
it, spending sorbent that sat idle while the front travelled; a simulated moving
bed runs adsorption continuously and counter-currently, and it is the industry
answer to the one separation that neither distillation nor crystallisation
handles cleanly, the splitting of two enantiomers that differ in nothing a
boiling point or a solubility can see. The vignette shows the same
equilibrium-and-rate physics of section~7.5 arranged into a continuous cycle,
and it closes the adsorption thread the third case study opened.

\subsection*{Why a moving bed, and why simulated}

A true counter-current adsorber would flow the solid downward against an upward
fluid, so fresh sorbent always meets the most-depleted fluid and loaded sorbent
leaves against the richest feed, the same counter-current advantage that made
the absorber and the extractor spend their agent efficiently. Moving a packed
solid bodily grinds the particles to dust and channels the flow, so the moving
bed is simulated instead: the packing stays fixed, joined in a loop of several
columns, and the four feed and product ports are advanced one column downstream
at fixed intervals, as figure~\ref{fig:vol04ch07:smb} draws it. Stepping the
ports against a circulating fluid is indistinguishable, over a full cycle, from
a solid creeping the other way, and it delivers the counter-current benefit with
no moving solid. The unit is the workhorse of preparative chromatography, from
fructose-glucose splitting in corn-syrup refining to the resolution of a
racemic drug into its active and inactive enantiomers on a chiral stationary
phase.

\input{volumes/04-energy/ch07-mass-transfer/figures/fig-smb.tex}

\subsection*{Sizing the zone flows}

The separation is set by four zone flow rates against the port-switch velocity,
and the sizing rule is a set of inequalities on the dimensionless flow-rate
ratios, one per zone. Define for each zone the ratio
\[
m_j = \frac{Q_j\,t^{*} - V\varepsilon}{V(1-\varepsilon)},
\]
with $Q_j$ the volumetric fluid flow in zone $j$, $t^{*}$ the switching
interval, $V$ the column volume, and $\varepsilon$ the bed voidage; $m_j$ is the
net fluid-to-solid flow ratio the zone presents. For a binary of a strongly
retained component with Henry constant $H_A$ and a weakly retained one with
$H_B < H_A$, the triangle-theory conditions for complete separation are
\[
m_1 > H_A,\quad H_B < m_2 < m_3 < H_A,\quad m_4 < H_B,
\]
which read directly: zone I must run fast enough to desorb even the strong
component and regenerate the sorbent, zone IV must run slow enough to adsorb
even the weak component and clean the recycle, and the two middle zones must
bracket the two Henry constants so the strong component is carried one way into
the extract and the weak one the other way into the raffinate. Take Henry
constants $H_A = 3.0$ and $H_B = 1.5$ for a chiral pair on the stationary
phase. A working design sets the middle zones at $m_2 = 1.8$ and $m_3 = 2.7$,
both inside the open interval $(1.5, 3.0)$, with $m_1 = 3.3$ above $H_A$ and
$m_4 = 1.2$ below $H_B$. The feed flow is the difference the two middle zones
can carry, $Q_F \propto (m_3 - m_2) = 0.9$ in the dimensionless units, so the
productivity rises with the gap between the two middle ratios, and that gap is
capped by the selectivity $H_A/H_B = 2.0$: a pair that differs little in
retention leaves only a narrow triangle of feasible operation and a low feed
throughput, the quantitative statement of why a poor chiral selectivity forces a
large, slow unit.

\subsection*{The verdict}

A four-zone simulated moving bed of eight columns, its ports stepped one column
per switching interval and its four zone flows placed inside the separation
triangle the two Henry constants define, resolves a racemic feed into its two
enantiomers continuously, at a sorbent and desorbent inventory a small fraction
of what a batch chromatograph would spend for the same purity and yield. The
design constraint with no analogue in the pressure-driven or distillation cases
is the separation triangle: the four flow-rate ratios must straddle the two
retention constants in the right order, and the feasible region shrinks toward a
point as the selectivity falls toward one, which is the continuous-adsorption
version of the vanishing relative volatility that stalled distillation at the
azeotrope. The unit is a distillation column turned inside out, with a solid
that only pretends to move and a counter-current contact built out of a clock
rather than a packing, and it is the standard answer wherever the components
differ only in how tightly a surface holds them
\cite{text:perry-chemical-engineers}.

\section{Failure: separation breakthroughs, fouling, channelling, equilibrium-stage failures}\pathtag{core}

Mass-transfer equipment fails in patterns analogous to heat-exchanger
failures (chapter~6) but with several distinctive modes.

\subsection{Breakthrough in adsorption beds}

A fixed-bed adsorber operates by loading the solid with solute from
the feed until the bed becomes saturated. The point at which the
outlet concentration begins to rise above the design limit is called
\engterm{breakthrough}, and the bed is then regenerated or replaced.
A common operational failure is to allow the bed to be used past
breakthrough, contaminating the downstream system with unprocessed
solute. The diagnostic is online monitoring of the bed-outlet
concentration; the engineering response is automated breakthrough
detection and switchover to a standby bed.

\subsection{Channelling in packed columns}

A packed column achieves separation through uniform contact between
the vapour and the liquid in the packing. \engterm{Channelling}
occurs when the liquid finds preferential paths down the column,
leaving regions of the packing dry; the apparent number of stages
drops dramatically. Causes: uneven liquid distribution at the top of
the column, packing settled in service to leave gaps, vapour
maldistribution at the bottom. The diagnostic is the column's
overall efficiency dropping below its commissioning value; the
remedy is mechanical (redistributor plates, repacking) or
operational (increasing the wetting rate).

\subsection{Equilibrium-stage assumption failures}

The design analysis treats each stage as reaching vapour-liquid
equilibrium. Real stages do not; tray efficiency is incorporated
through the Murphree efficiency $E_M$ (the fraction of the
equilibrium step actually achieved). When $E_M$ is lower than
designed, the column delivers fewer-than-designed effective stages.
Causes: foaming on the tray (gas trapped in liquid, reducing
contact); weeping (liquid dropping through the tray rather than
flowing across); entrainment (liquid droplets carried up by the
vapour, reducing separation). The diagnostic is column-product-
composition drift from design; the remedy depends on the specific
mechanism.

\subsection{Membrane fouling and breakthrough}

Membrane processes fail when the membrane surface is fouled (by
particulates, scaling, or biological growth) or when the membrane
itself loses integrity (a hole, a delamination). The two failures
have different signatures: fouling shows as declining flux at fixed
pressure; integrity loss shows as a sudden increase in permeate
contamination. Online monitoring of both flux and permeate quality
distinguishes them.

\subsection{Runaway reactions in mass-transfer-coupled systems}

Some processes combine mass transfer with chemical reaction in ways
that can run away. The Bhopal disaster of 1984 (registry:
\cite{acc:icmr-bhopal-2004}) involved a tank of methyl isocyanate
(MIC) into which water leaked, triggering a runaway exothermic
reaction. The chemical engineering of MIC storage is a combination
of heat transfer (the cooling system that should have removed the
reaction heat), mass transfer (the vapour vent that should have
released the gas to a scrubber, not to the atmosphere), and reaction
kinetics. The mass-transfer side of the failure: the vapour exited
the tank through a relief valve that vented to atmosphere because
the downstream scrubber was offline and the flare was out of
service; the engineered mass-transfer pathway from tank to safe
destruction had been broken.

The lessons by scale, from the registry entry: cooling capacity
must match reaction enthalpy generation (chapter~5 of this volume);
the vent path must deliver the gas to a safe sink; redundant
safeguards must be maintained in operating condition. The Bhopal
case appears in this chapter's failure section because it is the
canonical industrial example of a runaway reaction whose intended
mass-transfer mitigation was unavailable when needed.

\subsection{The diagnostic principle for mass-transfer equipment}

The maintenance discipline for separation equipment tracks:

\begin{itemize}[leftmargin=*]
  \item Product composition: drift from design indicates an upstream
    mass-transfer change.
  \item Pressure drop across the column or bed: rising indicates
    fouling or solids accumulation.
  \item Reflux ratio or stage temperatures: rising stage-by-stage
    temperatures (in a distillation column) suggest more reflux is
    needed to maintain the separation.
  \item Membrane flux: declining over time indicates fouling;
    sudden drops suggest membrane damage.
\end{itemize}

The diagnostic comes from comparing the operating-line analysis to
the actual stream measurements; the working position of any
separation engineer is to maintain instrumentation enough to make
the diagnosis routine.

\begin{principle}[The slowest film sets the separation rate]
A multi-phase mass-transfer process has resistances on both sides of
the interface, summing in series for the overall coefficient. The
phase carrying the dominant resistance sets the separation rate.
Identifying the dominant resistance is the first step in any design
optimisation: improvements to the dominant-resistance side reduce
the overall resistance; improvements to the minor-resistance side
have negligible effect. The same principle applies to heat
transfer (chapter~5) and to combined heat-and-mass-transfer
operations.
\end{principle}

\begin{estimation}
\textbf{Estimate the height of a packed-column rectifier to separate
a 50:50 ethanol-water mixture to $95\,\%$ ethanol overhead and $5\,\%$
ethanol in the bottoms.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
Ethanol forms an azeotrope with water at $89$ mol-\% ethanol
($95.5\,\%$ by mass) at $1\,\text{atm}$; a binary distillation cannot
exceed this composition. Take the design target as $89\,\%$ ethanol
overhead, $5\,\%$ ethanol in the bottoms.

The relative volatility $\alpha$ of ethanol to water varies from
$\sim 11$ at low ethanol concentration to $\sim 1.0$ at the
azeotrope. Take an average $\alpha \approx 2$.

The minimum number of stages from the Fenske equation:
\[
N_{\min} = \frac{\ln[(x_D/(1-x_D))((1-x_W)/x_W)]}{\ln \alpha} = \frac{
\ln[(0.89/0.11)(0.95/0.05)]}{\ln 2} = \frac{\ln 154}{\ln 2} = 7.3.
\]
Operating at $1.5 \times$ the minimum reflux requires perhaps $1.5$
to $2 \times$ the minimum stages, so $\sim 15$ equilibrium stages.

At an HETP of $0.3\,\text{m}$ for structured packing, the required
height is $\sim 4.5\,\text{m}$ of packed bed; with end zones and
distributor plates, the total column height is $\sim 6$ to
$8\,\text{m}$.

Published designs for ethanol-water rectifiers in beverage and
industrial alcohol production show columns in the $4$ to $10\,
\text{m}$ range for stage requirements of this magnitude
\cite{text:treybal-mass-transfer}. The estimate is in the
right range.

The postmortem: the dominant uncertainty is the average $\alpha$,
which varies substantially along the column. A more accurate
calculation would use the local $\alpha$ at each stage (and would
move toward operating well below the azeotrope to avoid the
asymptote in stage requirements). For overhead compositions
approaching the azeotrope (within $1$ percentage point), the stage
requirement diverges; this is why beverage distillation aims at the
azeotrope while industrial dehydrated ethanol uses an additional
unit operation (entrainer, molecular sieves) to cross the azeotrope.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Distil ethanol from a fermented sample and characterise
the column's HETP]
The reader sets up a small distillation column (a glass column with
random packing, or a commercial ``reflux'' still), distils a
fermented sample (beer, wine, or a sugar-yeast home-brew), and
characterises the column's separation performance.

\textbf{Track}. Household-scale experiment plus analysis.
\textbf{Hazard class}: low-to-moderate (alcohol is flammable;
distillation must occur outdoors or in a well-ventilated area; no
open flames). Compliance with local regulations on home distillation
is the reader's responsibility.

\textbf{Method}.

\begin{itemize}[leftmargin=*]
  \item Prepare a fermented sample with $\sim 8\,\%$ ethanol (a
    standard home brew or a commercial wine).
  \item Distil through the packed column at total reflux for a
    measured period, then collect overhead product at a controlled
    reflux ratio.
  \item Measure overhead composition by hydrometer (the ratio of
    overhead density to the known density-ethanol curves gives the
    mass fraction).
  \item Plot the cut versus boiler temperature; identify the
    transitions between low-boiling impurities (acetone, methanol
    in trace), ethanol cut, and water-rich tails.
\end{itemize}

\textbf{Analysis}.

\begin{itemize}[leftmargin=*]
  \item Compute the number of equilibrium stages achieved from the
    inlet and overhead compositions, using the Fenske equation for
    total reflux or McCabe-Thiele for finite reflux.
  \item Compute the HETP from the column packing height divided by
    the achieved equilibrium stages.
  \item Compare to literature HETP values for the column's packing
    type and operating conditions.
\end{itemize}

\textbf{Reflection ($1{,}500$ to $2{,}500$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
  \item Why is the achieved HETP higher or lower than literature?
    Identify the dominant source of efficiency loss.
  \item What is the relative volatility at the operating conditions,
    and how does it limit the achievable overhead composition?
  \item What would be needed to cross the ethanol-water azeotrope?
\end{itemize}

\textbf{Deliverable}. The data log, the analysis, the HETP
calculation, and the reflection.

\textbf{Time}. Three to six hours of distillation, two to three hours
of analysis, four to six hours of write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Water vapour diffuses through a stagnant air film of $5\,\si{\milli
\meter}$ thickness between a wet surface (saturation pressure
$2.34\,\text{kPa}$ at $20\,\degC$) and a dry-air bulk where the
partial pressure of water is $0.5\,\text{kPa}$. Compute the molar
flux of water vapour, given $D_{\text{water-air}} = 2.6 \times
10^{-5}\,\text{m}^2/\text{s}$, $p_{\text{total}} = 101\,\text{kPa}$.

\paragraph{Solution.}
\textit{Step 1, convert pressures to concentrations.} With $RT =
(8.314)(293) = 2436\,\text{J/mol}$, the surface concentration is
$c_{A,1} = 2340/2436 = 0.961\,\text{mol/m}^3$ and the bulk
$c_{A,2} = 500/2436 = 0.205\,\text{mol/m}^3$.

\textit{Step 2, bare Fick estimate.} For dilute water vapour,
$N_A = D(c_{A,1}-c_{A,2})/L = (2.6\times 10^{-5})(0.756)/(5\times
10^{-3}) = 3.9\times 10^{-3}\,\text{mol/m}^2\text{s}$.

\textit{Step 3, Stefan correction.} Air is stagnant, so water carries
the bulk flow. The mole fractions are $x_{A,1} = 2.34/101 = 0.0232$
and $x_{A,2} = 0.5/101 = 0.0050$; the log-mean inert fraction is
$x_{B,\text{lm}} = (0.995-0.977)/\ln(0.995/0.977) = 0.986$. Dividing
the bare flux by $x_{B,\text{lm}}$ raises it by about $1.4\,\%$ to
$N_A \approx 4.0\times 10^{-3}\,\text{mol/m}^2\text{s}$.

\textit{Step 4, comment.} The dilute approximation is good here
because the water mole fraction stays below $3\,\%$; the Stefan factor
only becomes a leading effect when the diffusing species is a large
fraction of the gas.
\end{exercise}

\begin{exercise}
Oxygen diffuses into still water from an air-water interface where
the dissolved oxygen concentration is $8\,\text{mg/L}$ (saturation at
$20\,\degC$). Find the time for the dissolved oxygen at a depth of
$5\,\si{\centi\meter}$ to reach half-saturation, treating the water
column as a semi-infinite medium with initial zero oxygen. Take
$D = 2.1 \times 10^{-9}\,\text{m}^2/\text{s}$.

\paragraph{Solution.}
\textit{Step 1, write the semi-infinite solution.} With surface
concentration $c_s$ and initial zero, the profile is $c/c_s =
\operatorname{erfc}\!\big(x/(2\sqrt{Dt})\big)$. Half-saturation means
$\operatorname{erfc}(\eta) = 0.5$, which gives the similarity variable
$\eta = x/(2\sqrt{Dt}) = 0.4769$.

\textit{Step 2, solve for time.} Rearranging, $t = x^2/(4 D \eta^2)$.
With $x = 0.05\,\si{\meter}$,
\[
t = \frac{(0.05)^2}{4(2.1\times 10^{-9})(0.4769)^2}
= 1.31\times 10^{6}\,\text{s} \approx 15\ \text{days}.
\]

\textit{Step 3, interpret.} Pure molecular diffusion is far too slow to
oxygenate standing water at this depth. Aeration ponds and lakes depend
on convective mixing (wind, thermal turnover, mechanical aerators), not
diffusion, to distribute dissolved oxygen.
\end{exercise}

\begin{exercise}
A binary distillation has $x_D = 0.95$, $x_W = 0.05$, feed at $z_F =
0.4$ at saturated liquid, relative volatility $\alpha = 2.5$.
Compute the minimum reflux ratio, the minimum number of stages, and
estimate the actual number of stages for $R = 1.3 R_{\min}$.

\paragraph{Solution.}
\textit{Step 1, minimum stages by Fenske.} For constant relative
volatility, $N_{\min} = \ln[(x_D/(1-x_D))((1-x_W)/x_W)]/\ln\alpha =
\ln[(0.95/0.05)(0.95/0.05)]/\ln 2.5 = \ln(361)/\ln 2.5 = 6.43$, so
about $6.4$ equilibrium stages at total reflux.

\textit{Step 2, minimum reflux.} For a saturated-liquid feed
($q = 1$) the q-line is vertical at $x = z_F = 0.4$; the pinch lies on
the equilibrium curve there, $y' = \alpha z_F/(1+(\alpha-1)z_F) =
(2.5)(0.4)/(1+1.5\cdot 0.4) = 1.0/1.6 = 0.625$. The minimum reflux
follows from the rectifying line passing through $(z_F, y')$ and
$(x_D, x_D)$:
\[
R_{\min} = \frac{x_D - y'}{y' - z_F} = \frac{0.95-0.625}{0.625-0.40}
= \frac{0.325}{0.225} = 1.44.
\]

\textit{Step 3, actual stages.} Operate at $R = 1.3 R_{\min} = 1.88$.
The Gilliland correlation maps $(R-R_{\min})/(R+1)$ to
$(N-N_{\min})/(N+1)$; with $X = (1.88-1.44)/2.88 = 0.153$, the
correlation gives $Y \approx 0.46$, so $N \approx (N_{\min}+Y)/(1-Y) =
(6.43+0.46)/0.54 \approx 12.8$, about $13$ equilibrium stages.

\textit{Step 4, sanity.} The stage count sits between the $6.4$ total-
reflux minimum and the divergence at $R_{\min}$, as expected at
$1.3\times$ minimum reflux.
\end{exercise}

\begin{exercise}
An absorber recovers $95\,\%$ of $\text{SO}_2$ from a flue gas at
$1\,\text{atm}$, $50\,\degC$, with $\text{SO}_2$ mole fraction
$0.005$. The solvent is dilute sodium hydroxide solution with $m =
0.1$. Use the Kremser equation to find the minimum liquid-to-gas
ratio and the number of stages at $1.5 \times$ minimum.

\paragraph{Solution.}
\textit{Step 1, minimum solvent rate.} The minimum liquid-to-gas ratio
corresponds to the exiting liquid leaving in equilibrium with the
entering gas, $(L/G)_{\min} = m\,(\text{fractional recovery}) = m\cdot
0.95 = 0.1\cdot 0.95 = 0.095$. Because $m$ is small the solvent demand
is modest; the gas is highly soluble in the caustic.

\textit{Step 2, operating solvent rate.} At $1.5\times$ minimum,
$L/G = 0.143$, so the absorption factor is $A = (L/G)/m = 0.143/0.1 =
1.43$.

\textit{Step 3, stages from Kremser.} For lean solvent the recovery is
$\phi = (A^{N+1}-A)/(A^{N+1}-1) = 0.95$. Solving,
$N = \ln[(1-\phi/A)/(1-\phi)]/\ln A - 1 =
\ln[(1-0.664)/0.05]/\ln 1.43 - 1 = \ln(6.72)/0.358 - 1 \approx 4.3$,
so about five theoretical stages.

\textit{Step 4, check.} With $A>1$ the absorber can in principle reach
any recovery given enough stages; the $1.5\times$ margin keeps the
stage count finite and the column economical.
\end{exercise}

\begin{exercise}
A reverse-osmosis seawater desalination plant operates at $7\,\text{MPa}$
feed pressure with feed osmotic pressure $2.7\,\text{MPa}$ and
permeate osmotic pressure negligible. Compute the permeate flux given
membrane permeability $A_w = 1.0 \times 10^{-12}\,\text{m}/(\text{s}
\cdot\text{Pa})$.

\paragraph{Solution.}
\textit{Step 1, net driving pressure.} The water flux is driven by the
applied pressure less the osmotic back-pressure, $\Delta p - \Delta\pi
= 7.0 - 2.7 = 4.3\,\text{MPa} = 4.3\times 10^{6}\,\text{Pa}$.

\textit{Step 2, flux.} $J_w = A_w(\Delta p - \Delta\pi) =
(1.0\times 10^{-12})(4.3\times 10^{6}) = 4.3\times 10^{-6}\,\text{m/s}$.

\textit{Step 3, convert to working units.} Multiplying by $3.6\times
10^{6}$ gives $J_w = 15.5\,\text{L}/(\text{m}^2\cdot\text{h})$, in the
typical seawater-RO range.

\textit{Step 4, caveat.} The calculation ignores concentration
polarisation, which raises the wall osmotic pressure above the bulk
value and reduces the real flux; the design driving force should use
the wall concentration, not the feed.
\end{exercise}

\begin{exercise}
A packed column for absorbing ammonia from air into water has
$H_G = 0.4\,\text{m}$ (height of gas-side transfer unit) and a
required $N_G = 5$ (number of transfer units). Compute the required
packing height.

\paragraph{Solution.}
\textit{Step 1, apply the transfer-unit decomposition.} The packed
height is the product of the height of a transfer unit and the number
of transfer units, $Z = H_G\,N_G$.

\textit{Step 2, evaluate.} $Z = (0.4\,\si{\meter})(5) = 2.0\,\si{
\meter}$ of packed bed.

\textit{Step 3, interpret.} $H_G$ packages the gas-side mass-transfer
coefficient and the packing area into a length; $N_G$ measures how many
e-folding driving-force reductions the separation demands. The split is
the mass-transfer analogue of HETP times stage count, but framed as a
rate process rather than an equilibrium-stage count. Add end zones and
a distributor and the column shell would be perhaps $3$ to
$3.5\,\si{\meter}$ tall.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive Fick's second law $\partial c_A/\partial t = D \laplacian c_A$
from the species mass balance and Fick's first law $\vect{J}_A = -D
\grad c_A$.

\paragraph{Solution.}
\textit{Step 1, write the species balance.} For a fixed control volume
with no convection and no reaction, conservation of $A$ states that the
accumulation equals the net diffusive inflow:
$\partial c_A/\partial t = -\divg \vect{J}_A$.

\textit{Step 2, substitute Fick's first law.} Insert $\vect{J}_A =
-D\grad c_A$:
$\partial c_A/\partial t = -\divg(-D\grad c_A) = \divg(D\grad c_A)$.

\textit{Step 3, constant diffusivity.} Taking $D$ uniform pulls it out
of the divergence, $\partial c_A/\partial t = D\,\divg(\grad c_A) =
D\laplacian c_A$, which is Fick's second law.

\textit{Step 4, note the analogy.} The same two steps, a conservation
law plus a linear flux-gradient relation, give the heat equation from
Fourier's law and the pressure-diffusion equation from Darcy's law; the
diffusion equation is the shared form of all three.
\end{exercise}

\begin{exercise}
Derive the Kremser equation for absorption with a linear equilibrium
relation $y^* = mx$ and a counter-current cascade.

\paragraph{Solution.}
\textit{Step 1, stage balance.} For dilute streams with constant flows
$L$ and $G$, a solute balance around stage $n$ gives $G y_{n+1} + L
x_{n-1} = G y_n + L x_n$. With equilibrium $y_n = m x_n$ on each stage,
eliminate $x$ to get a linear recurrence in the gas compositions.

\textit{Step 2, define the absorption factor.} Write $A = L/(mG)$. The
recurrence becomes $y_{n+1} - y_n = A(y_n - y_{n-1})$, a constant-ratio
relation: successive driving-force decrements form a geometric series
with ratio $A$.

\textit{Step 3, sum the cascade.} Summing the geometric series from the
top stage to the bottom over $N$ stages, with a lean entering solvent
($x_0 = 0$), the fraction of entering solute not absorbed is
$A^{N+1}-A$ over $A^{N+1}-1$ inverted, giving the recovery form
\[
\frac{y_{N+1}-y_1}{y_{N+1}-m x_0} = \frac{A^{N+1}-A}{A^{N+1}-1}.
\]

\textit{Step 4, limiting behaviour.} For $A = 1$ the series degenerates
and the recovery is $N/(N+1)$; for $A>1$ the recovery approaches $1$ as
$N\to\infty$; for $A<1$ it saturates below $1$ regardless of stage
count, the signal of an under-solvented absorber.
\end{exercise}

\begin{exercise}
Show that for diffusion through a stagnant gas film, the molar flux
of the diffusing species is $N_A = D_{AB}(c_{A,1} - c_{A,2})/(L (1 -
x_{A,\text{lm}}))$ rather than the simple Fick's-law expression.

\paragraph{Solution.}
\textit{Step 1, set up the flux with bulk flow.} Species $A$ diffuses
while $B$ is stagnant, so the total molar flux is $N_A$ with $N_B = 0$.
The combined convection-diffusion flux is
$N_A = -c D_{AB}\,dx_A/dz + x_A(N_A + N_B) = -c D_{AB}\,dx_A/dz +
x_A N_A$.

\textit{Step 2, isolate the gradient.} Rearranging,
$N_A(1 - x_A) = -c D_{AB}\,dx_A/dz$, hence
$N_A = -\dfrac{c D_{AB}}{1-x_A}\dfrac{dx_A}{dz}$.

\textit{Step 3, integrate across the film.} With $N_A$ constant at
steady state, separate variables and integrate from $z=0$ ($x_A =
x_{A,1}$) to $z=L$ ($x_A = x_{A,2}$):
\[
N_A L = c D_{AB}\,\ln\!\frac{1-x_{A,2}}{1-x_{A,1}}.
\]

\textit{Step 4, recast with the log mean.} Define the log-mean inert
fraction $x_{B,\text{lm}} = (x_{B,2}-x_{B,1})/\ln(x_{B,2}/x_{B,1})$
with $x_B = 1 - x_A$. Substituting and using $c(x_{A,1}-x_{A,2}) =
c_{A,1}-c_{A,2}$ gives
\[
N_A = \frac{D_{AB}(c_{A,1}-c_{A,2})}{L\,x_{B,\text{lm}}},
\]
the Stefan-corrected flux. The factor $1/x_{B,\text{lm}}$ exceeds unity
and reduces to $1$ in the dilute limit, recovering Fick's law.
\end{exercise}

\begin{exercise}
Derive the McCabe-Thiele operating-line equations for the rectifying
and stripping sections of a binary distillation column.

\paragraph{Solution.}
\textit{Step 1, rectifying envelope.} Draw a balance around the top of
the column, enclosing the condenser and all stages above stage $n$.
Constant molal overflow makes the vapour $V$ and liquid $L$ flows
uniform in the section. The light-component balance is $V y_{n+1} =
L x_n + D x_D$.

\textit{Step 2, rectifying line.} Divide by $V$ and use $L/V = R/(R+1)$
and $D/V = 1/(R+1)$ for reflux ratio $R = L/D$:
\[
y_{n+1} = \frac{R}{R+1}\,x_n + \frac{x_D}{R+1}.
\]
This is a straight line of slope $R/(R+1)$ through $(x_D, x_D)$.

\textit{Step 3, stripping envelope.} Balance around the bottom,
enclosing the reboiler and all stages below stage $m$, with stripping-
section flows $L'$ and $V'$: $L' x_{m-1} = V' y_m + W x_W$, giving
\[
y_m = \frac{L'}{V'}\,x_{m-1} - \frac{W x_W}{V'},
\]
a straight line of slope $L'/V'$ through $(x_W, x_W)$.

\textit{Step 4, feed line.} A balance around the feed stage relates the
two sections through the feed quality $q$; the q-line $y = q/(q-1)\,x -
z_F/(q-1)$ is where the two operating lines intersect. The three lines
plus the equilibrium curve close the construction.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the rate of evaporation from a swimming pool of $50\,\text{m}
^2$ surface area in $25\,\degC$ ambient air with $50\,\%$ relative
humidity and a $2\,\text{m/s}$ wind. Use a wind-driven mass-transfer
coefficient correlation.

\paragraph{Solution.}
\textit{Step 1, driving concentration.} The saturation vapour pressure
of water at $25\,\degC$ is $3.17\,\text{kPa}$; the air at $50\,\%$
humidity carries $1.58\,\text{kPa}$. The surface-to-bulk concentration
difference is $\Delta c = (3170-1580)/(RT) = 1590/2479 =
0.64\,\text{mol/m}^3$, with $RT = (8.314)(298)$.

\textit{Step 2, mass-transfer coefficient.} A working pool correlation
gives $k_c \sim 10^{-2}\,\text{m/s}$ for a $2\,\text{m/s}$ wind over an
open-water surface (equivalently $h_m \approx 0.01\,\text{m/s}$ from
the Chilton-Colburn analogy applied to the convective heat-transfer
coefficient over the pool).

\textit{Step 3, evaporation rate.} The molar flux is $N_A = k_c\Delta c
= (10^{-2})(0.64) = 6.4\times 10^{-3}\,\text{mol/m}^2\text{s}$. Over
$50\,\text{m}^2$, $\dot n = 0.32\,\text{mol/s}$; multiplying by the
$0.018\,\text{kg/mol}$ molar mass gives $\dot m \approx 5.8\times
10^{-3}\,\text{kg/s} \approx 21\,\text{kg/h}$, roughly $20\,\text{L/h}$.

\textit{Step 4, sanity.} A residential pool losing tens of litres an
hour on a breezy day matches everyday experience; the dominant
uncertainty is $k_c$, which scales with wind speed roughly to the
$0.8$ power.
\end{exercise}

\begin{exercise}
Estimate the number of theoretical stages required to remove $90\,\%$
of acetone from a water stream by extraction with toluene
($K = y/x = 1.5$) at a solvent-to-feed ratio of $2$.

\paragraph{Solution.}
\textit{Step 1, extraction factor.} The extraction analogue of the
absorption factor is $E = K\,(S/F) = 1.5\times 2 = 3.0$, with $S/F$ the
solvent-to-feed ratio.

\textit{Step 2, apply the Kremser form.} For dilute counter-current
extraction with fresh solvent, the fraction of solute remaining in the
raffinate after $N$ stages is $(E-1)/(E^{N+1}-1)$. A $90\,\%$ recovery
leaves a fraction $0.10$:
\[
\frac{E-1}{E^{N+1}-1} = 0.10 \ \Rightarrow\ E^{N+1} = 1 + \frac{2}{0.10}
= 21 \ \Rightarrow\ N+1 = \frac{\ln 21}{\ln 3} = 2.77.
\]

\textit{Step 3, round up.} $N \approx 1.8$, so two theoretical stages
deliver slightly more than $90\,\%$ recovery; a single stage would fall
short.

\textit{Step 4, interpret.} With $E = 3$ the extraction is comfortable;
the favourable partition and the doubled solvent rate keep the stage
count low.
\end{exercise}

\begin{exercise}
Estimate the membrane area required for a $1000\,\text{m}^3/\text{day}$
seawater RO desalination plant, given typical permeate flux of $20\,
\text{L}/(\text{m}^2\cdot\text{h})$.

\paragraph{Solution.}
\textit{Step 1, convert the production rate.} $1000\,\text{m}^3/
\text{day} = 1.0\times 10^{6}\,\text{L}/\text{day} = 4.17\times 10^{4}
\,\text{L/h}$.

\textit{Step 2, divide by the flux.} Membrane area $= \dot V/J =
(4.17\times 10^{4})/20 = 2.1\times 10^{3}\,\text{m}^2$.

\textit{Step 3, translate to modules.} At roughly $40\,\text{m}^2$ of
membrane per standard spiral-wound element, the plant needs about
$50$ elements, packaged a handful per pressure vessel.

\textit{Step 4, caveat.} The flux figure already bakes in a fouling
margin; sizing on the clean-membrane flux would underprovision area and
force the plant to run at higher pressure as the membranes age.
\end{exercise}

\begin{exercise}
Estimate the drying time for a wet sand bed of $1\,\text{m}^2$ surface
area and $10\,\si{\milli\meter}$ initial moisture layer thickness in
$25\,\degC$ air at $30\,\%$ relative humidity with a $1\,\text{m/s}$
wind.

\paragraph{Solution.}
\textit{Step 1, water to remove.} A $10\,\si{\milli\meter}$ layer over
$1\,\text{m}^2$ is $0.010\,\text{m}^3 = 10\,\text{kg}$ of water,
equivalently $556\,\text{mol}$.

\textit{Step 2, constant-rate flux.} The surface holds near the
wet-bulb temperature; take the saturation deficit from the surface
saturation at the wet-bulb (about $19\,\degC$, $p_{\text{sat}}\approx
2.2\,\text{kPa}$) to the bulk ($0.3\times 3.17 = 0.95\,\text{kPa}$),
$\Delta c = (2200-950)/2479 = 0.50\,\text{mol/m}^3$. With a
$1\,\text{m/s}$ wind, $k_c \sim 7\times 10^{-3}\,\text{m/s}$, so
$N_A = k_c\Delta c = 3.5\times 10^{-3}\,\text{mol/m}^2\text{s}$.

\textit{Step 3, drying time.} The molar removal rate over $1\,\text{m}^2$
is $3.5\times 10^{-3}\,\text{mol/s}$; the constant-rate period to clear
the free water is $t = 556/(3.5\times 10^{-3}) = 1.6\times 10^{5}
\,\text{s} \approx 44\,\text{h}$.

\textit{Step 4, caveat.} This is the constant-rate estimate only; once
the sand passes its critical moisture the falling-rate period adds
substantial time, so the figure is a lower bound on total drying time.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a binary McCabe-Thiele solver: given the equilibrium curve
(via $\alpha$ or a tabulated $y$-$x$ relation), feed conditions,
distillate and bottoms specifications, and reflux ratio, return the
number of stages and the feed-tray location.

\paragraph{Solution.}
\textit{Step 1, encode the equilibrium and operating lines.} For
constant relative volatility use $y = \alpha x/(1+(\alpha-1)x)$ and its
inverse $x = y/(\alpha-(\alpha-1)y)$. The rectifying line is $y =
R/(R+1)\,x + x_D/(R+1)$; the stripping line passes through $(x_W, x_W)$
and the q-line pinch.

\textit{Step 2, step the staircase.} Start at $(x_D, x_D)$. Step
horizontally to the equilibrium curve (apply the inverse), then
vertically to whichever operating line applies at that $x$. Each
horizontal step is one equilibrium stage; switch from the rectifying to
the stripping line once $x$ falls below the feed pinch, recording that
step as the feed tray.

\textit{Step 3, terminate and verify.} Stop when $x \le x_W$; the step
count is the stage requirement. The reference implementation is
\texttt{code/mccabe\_thiele.py}; verify it against the chapter
estimation (a $50:50$ ethanol-water cut needs of order $15$ stages) and
against the limiting cases of total reflux (operating line on the
diagonal) and minimum reflux (operating line touching the equilibrium
curve, stages diverging).

\textit{Step 4, extend.} For a tabulated $y$-$x$ relation, replace the
constant-$\alpha$ functions with interpolation on
\texttt{data/ethanol\_water\_vle.csv}; the stepping logic is unchanged.
\end{exercise}

\begin{exercise}
Simulate transient diffusion of solute from a slab into a stirred
solution. Implement a finite-difference solver for the diffusion
equation in the slab coupled to a well-mixed liquid-side boundary
condition. Verify against the analytical solution for the early-time
limit.

\paragraph{Solution.}
\textit{Step 1, discretise.} Place $n_x$ nodes across the slab and
apply the explicit FTCS update $c_i^{k+1} = c_i^k + r(c_{i+1}^k -
2c_i^k + c_{i-1}^k)$ with $r = D\,\Delta t/\Delta x^2$. Keep $r \le
0.5$ for stability; the reference uses $r = 0.4$.

\textit{Step 2, set the boundary conditions.} Fix the surface node at
the equilibrium concentration with the stirred liquid, and impose a
no-flux (symmetry) condition at the far face by mirroring the interior
node. Couple the liquid by a solute balance if its concentration is
allowed to rise.

\textit{Step 3, verify early time.} For times short enough that the
front has not reached the far face, the slab behaves as semi-infinite
and the numerical surface-region profile must match
$\operatorname{erfc}(x/(2\sqrt{Dt}))$. The penetration depth should
track $\sqrt{Dt}$; \texttt{code/transient\_diffusion.py} prints both
for comparison.

\textit{Step 4, late time.} As the front reaches the symmetry plane the
profile flattens toward the equilibrium concentration; check that the
total solute uptake approaches the equilibrium loading and conserves
mass to the tolerance set by $\Delta x$ and $\Delta t$.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a packed absorption column to remove $99\,\%$ of $\text{H}_2
\text{S}$ from a natural-gas stream at $1\,\text{MPa}$ and $40\,\degC$
using monoethanolamine solution. Specify the packing type, the
column diameter (from flooding considerations), and the packing
height.

\paragraph{Solution.}
\textit{Step 1, choose the contactor.} Acid-gas removal into amine is
gas-side and liquid-side coupled with a reactive solvent; a counter-
current packed column with structured or large random packing (for low
pressure drop at the high gas throughput of a gas plant) is the
standard choice.

\textit{Step 2, size the diameter from flooding.} Use a generalised
pressure-drop correlation: pick the gas and liquid mass velocities,
read the flooding gas velocity, then operate at $60$ to $70\,\%$ of
flooding. The column area is the volumetric gas rate divided by the
operating superficial velocity; the diameter follows from the area.

\textit{Step 3, set the height.} The reactive amine drives $m$ small,
so gas-side resistance governs. Compute $N_{OG}$ from the required
$99\,\%$ recovery with the operating and equilibrium lines, estimate
$H_{OG}$ from a packing mass-transfer correlation, and take $Z =
H_{OG}N_{OG}$; a deep-cut acid-gas absorber typically lands at several
metres of packing.

\textit{Step 4, state assumptions.} Report the amine loading limit
(to avoid corrosion), the solvent circulation rate, and the regenerator
duty; the absorber sizing is incomplete without the paired stripper
that returns lean amine.
\end{exercise}

\begin{exercise}
Specify a freeze-dryer design for $100\,\text{kg/day}$ of
pharmaceutical product with $90\,\%$ initial water content and
$2\,\%$ final water content. Include the sublimation-rate analysis
and the heat-supply requirement.

\paragraph{Solution.}
\textit{Step 1, water load.} From $90\,\%$ to $2\,\%$ moisture, the
removed water is $100\,\text{kg/day}\times(0.90-0.02)/(1-0.02)$ of dry
basis, about $88\,\text{kg/day}$ of water, dominated by the primary-
drying sublimation step.

\textit{Step 2, sublimation rate.} Spread the load over the working
drying cycle (say $20\,\text{h}$), giving $\sim 4.4\,\text{kg/h}$ of
ice to sublime. The vapour flux is set by the resistance of the dried
cake and the chamber-to-condenser pressure difference; at the low
chamber pressure ($\sim 10$ to $30\,\text{Pa}$) the specific vapour
volume is large, which sizes the vapour duct and condenser.

\textit{Step 3, heat supply.} Each kilogram of ice needs the heat of
sublimation, $\sim 2.84\,\text{MJ/kg}$. The shelf-heat duty is
$4.4\,\text{kg/h}\times 2.84\,\text{MJ/kg} \approx 12.5\,\text{MJ/h}
\approx 3.5\,\text{kW}$, delivered by conduction through the shelves
without melting the product (shelf temperature limited by the collapse
temperature of the cake).

\textit{Step 4, condenser.} The condenser must trap all $88\,\text{kg/
day}$ of vapour as ice at a temperature well below the chamber, and its
refrigeration duty roughly matches the shelf heat plus the chamber
losses.
\end{exercise}

\begin{exercise}
Design a multi-stage spray dryer for producing $1000\,\text{kg/day}$
of milk powder from feed milk at $4\,\degC$ to product at $3\,\%$
moisture. Specify the air flow, the inlet and outlet air
temperatures, and the chamber volume.

\paragraph{Solution.}
\textit{Step 1, water load.} Milk feed is roughly $88\,\%$ water; to
produce $1000\,\text{kg/day}$ of $3\,\%$-moisture powder the feed
solids are $\sim 970\,\text{kg/day}$, requiring feed of $\sim
8000\,\text{kg/day}$ and evaporation of $\sim 7000\,\text{kg/day}$ of
water, about $290\,\text{kg/h}$.

\textit{Step 2, air temperatures.} Spray dryers run a hot inlet
($\sim 180$ to $200\,\degC$) and a moderate outlet ($\sim 85$ to
$95\,\degC$) chosen to keep the powder below its scorch limit while
leaving the air able to carry the evaporated water.

\textit{Step 3, air flow.} The drying air supplies the latent heat. An
energy balance, $\dot m_{\text{air}}c_{p}\Delta T = \dot m_{\text{water}}
h_{fg}$ with $\Delta T \approx 100\,\degC$, $c_p \approx 1.05\,\text{kJ/
kg K}$, $h_{fg}\approx 2.3\,\text{MJ/kg}$, gives $\dot m_{\text{air}}
\approx (290/3600)(2.3\times 10^{6})/(1050\times 100) \approx
1.8\,\text{kg/s}$ of air.

\textit{Step 4, chamber volume.} Size the chamber from the droplet
residence time needed for drying (a few seconds for fine atomised
droplets) times the air volumetric flow; at $\sim 1.5\,\text{m}^3/
\text{kg}$ hot-air specific volume and a few seconds' residence, the
chamber is of order $10\,\text{m}^3$, with the exact value set by the
atomiser droplet-size distribution.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A distillation column's overhead composition has drifted from
$98\,\%$ to $94\,\%$ ethanol over six months. The reflux ratio and
boiler duty are unchanged. Identify the most probable cause
(foaming, weeping, packing settling, fouling on internals) and the
diagnostic measurement.

\paragraph{Solution.}
\textit{Step 1, frame the symptom.} A slow, steady efficiency loss at
fixed reflux and duty points to a gradual mechanical degradation of
contact, not a sudden upset. The candidates are packing settling
(channelling) or fouling on the internals; foaming and weeping tend to
appear with rate or composition changes rather than over months.

\textit{Step 2, rank causes.} Over six months at unchanged operating
conditions, packing settling that opens channels and fouling that
blinds packing area are the most probable; both lower the effective
stage count and degrade the overhead.

\textit{Step 3, diagnostic measurement.} Measure the pressure drop
across the packed section and compare to the commissioning value.
Fouling and solids accumulation raise $\Delta p$; channelling from
settled packing tends to lower $\Delta p$ as gas finds open paths. A
gamma-scan of the column localises the maldistributed or fouled zone.

\textit{Step 4, response.} If $\Delta p$ has risen, plan a cleaning
shutdown; if it has fallen, install a redistributor or repack, and
restore the wetting rate to spread the liquid evenly.
\end{exercise}

\begin{exercise}
A reverse-osmosis plant's permeate flux has fallen by $25\,\%$ over
three months at fixed feed pressure. Identify the most probable
fouling mechanism (scaling, particulate, biofouling) and the
diagnostic protocol.

\paragraph{Solution.}
\textit{Step 1, read the signature.} A gradual flux decline at fixed
pressure is the classic fouling signature; the three mechanisms differ
in their fingerprints. Scaling raises the differential pressure modestly
and worsens with recovery; particulate fouling spikes the lead-element
differential pressure; biofouling shows a creeping differential-
pressure rise and a microbiological count in the feed.

\textit{Step 2, locate the foulant.} Profile the per-stage differential
pressure and the per-stage salt passage. Front-end differential-
pressure rise points to particulate or biofouling; back-end loss points
to scaling, which concentrates where the brine is most saturated.

\textit{Step 3, confirm by autopsy and chemistry.} Pull a representative
element, weigh and inspect the foulant, and run a chemical analysis
(carbonate or sulphate scale, organic biofilm, or inorganic colloids).
Check the feed silt density index and the antiscalant dose log.

\textit{Step 4, response.} Match the clean to the foulant: acid clean
for carbonate scale, alkaline or enzymatic clean for biofilm, improved
pretreatment (filtration, antiscalant) for particulates and scaling.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Re-read the ICMR Bhopal report \cite{acc:icmr-bhopal-2004}. Identify
the role of the vapour-vent path (the mass-transfer route from tank
to atmosphere) in the disaster. State the engineered safeguards
that should have intercepted the vapour and explain why each failed.

\paragraph{Solution.}
\textit{Step 1, trace the intended path.} The engineered mass-transfer
route carried vapour from the storage tank to a relief device, then to
a caustic scrubber to neutralise it, and finally to a flare to burn any
residual. Each stage is a separation or destruction step on the vapour
stream.

\textit{Step 2, identify the failed safeguards.} The registry entry
\cite{acc:icmr-bhopal-2004} records that the refrigeration that should
have kept the tank cold was off, the scrubber was not in service, and
the flare was out of commission; the vent-gas path therefore terminated
at the atmosphere rather than at a safe sink.

\textit{Step 3, explain each failure.} Cooling off allowed the
water-triggered exotherm to raise tank pressure unchecked; the offline
scrubber removed the chemical-destruction stage; the out-of-service
flare removed the last barrier. The series of safeguards had no
remaining intact link.

\textit{Step 4, state the lesson.} A vent path is only a safeguard while
every stage in it is maintained operational; redundancy on paper is
worthless if the redundant equipment is not kept in working condition.
The mass-transfer route from hazard to safe destruction must be treated
as a continuously verified system, not a one-time installation.
\end{exercise}

\begin{exercise}
A pharmaceutical fermenter produces a batch of antibiotic with $40\,\%$
lower titre than the target. The temperature, pH, and dissolved
oxygen log normally. Identify the most probable mass-transfer-related
cause (oxygen-transfer limitation, substrate-feed-rate limitation,
cell-wall-permeability change) and the diagnostic.

\paragraph{Solution.}
\textit{Step 1, read the logged variables.} Temperature, pH, and
dissolved oxygen all logged normal, so the gross environment was in
spec. A normal dissolved-oxygen trace does not by itself clear oxygen
transfer, because the probe reads the bulk liquid, not the
oxygen-transfer rate at peak demand.

\textit{Step 2, rank causes.} The most probable mass-transfer cause is
an oxygen-transfer-rate limitation during peak growth, where the
volumetric coefficient $k_La$ cannot keep the cells supplied even
though the bulk dissolved oxygen has not yet crashed. Substrate-feed
limitation would usually show in the feed log; a cell-wall change is a
biological, not a transport, fault.

\textit{Step 3, diagnostic.} Measure $k_La$ directly by a dynamic
gassing-out test, and check the off-gas oxygen uptake rate against the
expected cellular demand. A $k_La$ below design, or an oxygen uptake
rate that plateaus while demand should still be rising, confirms a
transfer limitation.

\textit{Step 4, response.} Raise $k_La$ by increasing agitation,
aeration, or back-pressure, or reduce peak demand by trimming the feed
profile; re-run with the same logged set points to confirm the titre
recovers.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer argues that ``mass transfer is just diffusion.''
Write a one-paragraph reply that distinguishes molecular diffusion,
convective mass transfer, and the role of phase equilibrium in
separation.

\paragraph{Solution.}
\textit{Model reply.} Molecular diffusion is species transport down a
concentration gradient in a quiescent medium, governed by Fick's law
and the diffusivity $D$; it is slow, with liquid-phase time scales of
days over centimetres. Convective mass transfer is transport between a
surface and a moving fluid, packaged into a coefficient $k_c$ that
already contains the flow field, and it is orders of magnitude faster
because the boundary layer, not the bulk, carries the gradient. Phase
equilibrium is the third element and the one that makes separation
possible: a species partitions between two phases according to an
equilibrium relation (Henry's law, a partition coefficient, a
vapour-liquid equilibrium curve), and the achievable separation is set
by that equilibrium, while diffusion and convection only govern how
fast the system approaches it. Calling all of this ``just diffusion''
discards the two elements that do the engineering work.
\end{exercise}

\begin{exercise}
A facilities manager proposes to switch a distillation column from
tray internals to structured packing for ``efficiency.'' Write a
one-paragraph reply identifying the conditions under which the switch
pays off and the conditions under which it does not.

\paragraph{Solution.}
\textit{Model reply.} Structured packing pays off where pressure drop
matters and the service is clean: vacuum distillation, heat-sensitive
feeds that cannot tolerate the higher bottoms temperature a tray
column's pressure drop imposes, and revamps that need more separation in
an existing shell, where the lower HETP of structured packing buys
stages without added height. The switch does not pay off, and can
backfire, where the feed fouls or carries solids, because packing
channels and fouls in ways that are hard to clean, whereas trays
tolerate dirt and can be washed; it also disappoints where liquid
distribution is poor, since structured packing depends on uniform
irrigation and loses efficiency badly when maldistributed. The decision
turns on pressure-drop sensitivity and fouling tendency, not on a
generic claim of ``efficiency''; ask what the column's binding
constraint is before recommending the change.
\end{exercise}
