\chapter{Index of Archetypes}
\label{app:archetypes}

\noindent
Engineering is organised, across the twelve books, around ten recurring problem archetypes. The archetype is a formal environment (\texttt{\textbackslash begin\{archetype\}}) introduced in the book where it first appears and re-invoked in every later book that needs it. This appendix collects every invocation.

\TODO{Final population happens once the prose is written. Initial pass below establishes the structure.}

\section*{Balance}
\TODO{Cross-references to balance archetypes across Books I (dimensional balance), II (system equations), III (force), IV (energy), V (mass in chemistry), VII (information), IX (system equilibria), XII (resource accounting).}

\section*{Transport}
\TODO{Cross-references across Books III (momentum), IV (heat), V (mass transfer), VI (metabolic flux), VII (packets, queues), IX (network flow), XII (logistics).}

\section*{Stability}
\TODO{Cross-references across Books III (vibration), IV (thermodynamic equilibrium), V (phase), VI (homeostasis), VII (numerical), VIII (mechanical), IX (control), X (runaway).}

\section*{Optimisation}
\TODO{Cross-references across Books II (calculus of variations), III (least action), IV (efficiency), VII (algorithms), VIII (machine design), IX (operations), XI (design).}

\section*{Scaling}
\TODO{Cross-references across Books I (orders of magnitude), III (similitude), IV (heat-transfer scaling), V (microstructure to bulk), VI (cell to organism), VII (algorithmic), VIII (manufacturing), XII (planetary).}

\section*{Failure}
\TODO{Cross-references across Books I (measurement error), III (yield), IV (heat death), V (corrosion), VI (cell death), VII (bugs), VIII (wear), IX (cascading), X (everything).}

\section*{Control}
\TODO{Cross-references across Books III (statics as zero-order), IV (thermostats), VI (homeostasis), VII (feedback algorithms), VIII (servos), IX (whole-systems), XII (governance).}

\section*{Interface}
\TODO{Cross-references across Books I (sensor to system), III (boundary conditions), IV (heat exchange), V (surfaces), VI (membranes), VII (APIs), VIII (couplings), IX (subsystems), XI (human-machine).}

\section*{Uncertainty}
\TODO{Cross-references across Books I (error propagation), II (probability), III (random loads), IV (statistical mechanics), VII (Shannon entropy), IX (Kalman, robust control), X (rare events), XI (design margins).}

\section*{Ethics}
\TODO{Cross-references across Books V (environmental harm), VI (bioethics), VII (algorithmic harm), IX (autonomous systems), X (organisational), XI (under risk), XII (primarily).}
