\chapter{Polymers}
\label{vol05:ch06}

\epigraph{For lack of any better demonstration we must conclude that
the high-molecular-weight compounds, which we are concerned with here,
are not colloidal aggregates of small molecules but are true compounds
of high molecular weight built up by ordinary chemical bonds.}{Hermann
Staudinger, 1920, the founding statement of the macromolecular
hypothesis \cite[pp.~1073--1085]{hist:staudinger1920}.}

\chapmeta{Half-life: foundational for polymer physics; medium-life for commercial polymer grades and additive packages. Archetypes: scaling (chain to bulk), failure. Exercise target: 30. Project track: physical characterisation of household polymer objects. Hazard class: low (visual inspection, hand-pulling of samples, no heated processing). Reader path default: standard, with sections explicitly tagged. Named cases: none required; the polymer failure modes are described by mechanism.}

\noindent
A polymer is a long covalent molecule built by repetition of a small
chemical unit (the \engterm{monomer}). The macroscopic properties of
a polymer solid are set by the chain length, the chemical identity of
the monomer, the architecture of branching and crosslinking, and the
secondary bonding between chains. The chapter develops the polymer
chemistry that produces commercial materials, the thermal transitions
($T_g$ and $T_m$) that separate the working temperature ranges of the
polymer classes, the viscoelastic mechanical behaviour that
distinguishes polymers from metals and ceramics, the seven or so
engineering polymers that cover most application space (polyethylene,
polypropylene, polystyrene, ABS, polycarbonate, PEEK, polytetra-
fluoroethylene), the processing routes (extrusion, injection moulding,
additive manufacturing) by which polymer parts are made, and the
degradation modes (oxidation, photo-oxidation, hydrolysis, stress
cracking, creep) that limit polymer service life. The chapter is the
working-engineer companion to Chapter~1's bond-level framework and to
the materials-selection chapter (Chapter~10) of this volume.

\section{Polymer chemistry: chains, crosslinks, networks}\pathtag{core}

A \engterm{polymer} is a molecule with degree of polymerisation
$N \gg 1$, where $N$ is the number of repeat units. Engineering
polymers have $N$ in the range $100$ to $100\,000$; the resulting
molecular masses are $10^4$ to $10^7\,\text{g}/\text{mol}$, far larger
than the small molecules of organic chemistry. The transition from
small molecule to polymer is not sharp; oligomers ($N \sim 2$ to
$20$) behave more like small molecules, and full polymers behave
qualitatively differently from any small molecule with the same
repeat unit.

\engterm{Synthesis}. Polymers are produced by two main reaction
classes:

\begin{itemize}[leftmargin=*]
\item \engterm{Addition (chain-growth) polymerisation}: monomers with
  a reactive double bond (the vinyl family) add to a growing chain,
  each addition leaving the chain end still active. Polyethylene
  ($\text{CH}_2=\text{CH}_2 \to (\text{CH}_2-\text{CH}_2)_N$),
  polypropylene, polystyrene, polyvinyl chloride, and polymethyl
  methacrylate are produced by addition polymerisation. The reaction
  mechanism may be free-radical (initiated by peroxides or by light),
  cationic, anionic, or organometallic-catalysed (the Ziegler-Natta
  and metallocene catalysts that enabled the industrial production
  of polyolefins).
\item \engterm{Condensation (step-growth) polymerisation}: monomers
  with reactive end groups react in pairs, releasing a small-molecule
  by-product (often water). Polyesters, polyamides (nylons),
  polyurethanes, and polycarbonates are produced by condensation
  polymerisation. The reaction is slower than addition polymerisation
  and produces a more controlled molecular weight, set by the
  stoichiometry of the end groups.
\end{itemize}

\engterm{Molecular weight distribution}. Real polymers are mixtures of
chains with a distribution of lengths. The \engterm{number-average
molecular weight} is $\bar{M}_n = \sum n_i M_i / \sum n_i$, where
$n_i$ is the number of molecules of mass $M_i$; the
\engterm{weight-average molecular weight} is $\bar{M}_w = \sum n_i
M_i^2 / \sum n_i M_i$. The ratio $\text{PDI} = \bar{M}_w/\bar{M}_n$ is
the \engterm{polydispersity index}, with $\text{PDI} = 1$ for a
hypothetical monodisperse polymer and values from $\sim 1.05$ (anionic
polymerisation of narrow-distribution standards) to $\sim 30$ (some
high-density polyethylene grades) in commercial products. The
mechanical properties depend more strongly on $\bar{M}_w$ than on
$\bar{M}_n$ because the longer chains dominate the entanglement
network that gives polymers their characteristic toughness.

\engterm{Chain architecture}. The arrangement of monomer units along
the chain and the chain's branching topology:

\begin{itemize}[leftmargin=*]
\item \engterm{Linear} chains: a single long backbone. The structure
  of high-density polyethylene (HDPE), of linear polyamides, of
  most engineering thermoplastics.
\item \engterm{Branched} chains: a backbone with side chains of
  varying length and frequency. Low-density polyethylene (LDPE)
  has many short and long branches; the branching prevents close
  packing of chains and reduces density.
\item \engterm{Crosslinked} networks: chains tied together by covalent
  bonds. Lightly crosslinked networks (a few crosslinks per thousand
  monomer units) produce elastomers; densely crosslinked networks
  produce rigid thermosets.
\item \engterm{Block} copolymers: two or more chemically distinct
  segments along the same chain (an A-B diblock, an A-B-A triblock).
  Used to combine the properties of two polymers (the styrene-butadiene-
  styrene triblock is the model toughened elastomer; the rigid
  styrene end-blocks aggregate into hard domains that act as physical
  crosslinks in a soft butadiene matrix).
\item \engterm{Graft} copolymers: a backbone of one polymer with side
  chains of another. ABS plastic (acrylonitrile-butadiene-styrene)
  uses graft architecture to combine impact resistance from
  butadiene-rich domains with stiffness from styrene-acrylonitrile
  copolymer matrix.
\end{itemize}

\engterm{Stereochemistry (tacticity)}. The arrangement of side groups
along the chain in vinyl polymers can be:

\begin{itemize}[leftmargin=*]
\item \engterm{Isotactic}: side groups all on the same side of the
  backbone. Isotactic polypropylene is the commercial grade; the
  regular structure allows crystallisation.
\item \engterm{Syndiotactic}: side groups alternating on opposite
  sides. Crystallises with a different unit cell than isotactic;
  used for specialty grades.
\item \engterm{Atactic}: random arrangement. Atactic polypropylene
  is rubbery and non-crystalline at room temperature; commercial
  polypropylene contains $\sim 95\,\%$ isotactic chains.
\end{itemize}

The tacticity is controlled by the polymerisation catalyst.
Ziegler-Natta catalysts (developed by Karl Ziegler and Giulio Natta
in the 1950s, Nobel Prize 1963) produce isotactic polyolefins;
metallocene catalysts (commercialised in the 1990s) produce more
narrowly distributed and more uniformly tactic polymers.

\section{Thermoplastics vs thermosets vs elastomers}\pathtag{core}

The three macroscopic classes of polymers differ in how their chains
relate to each other above the melting or glass-transition temperature.

\engterm{Thermoplastics}. Long linear or branched chains held together
by secondary bonds (van der Waals, hydrogen, dipole-dipole). At room
temperature the secondary bonds give the bulk material its mechanical
properties; above the melting point (for semicrystalline polymers) or
above the glass transition (for amorphous polymers), the chains can
slip past each other and the material flows. Thermoplastics can be
melted, processed, and resolidified repeatedly without chemical
change, which makes them recyclable and easy to mould. The
engineering thermoplastics include polyethylene, polypropylene,
polystyrene, polyvinyl chloride, polyethylene terephthalate, nylon,
polycarbonate, ABS, and PEEK.

\engterm{Thermosets}. Covalently crosslinked networks formed by
chemical reaction during processing. Once cured, the network cannot
be melted; heating produces only chemical decomposition. The
crosslinks make the material harder and stiffer than the equivalent
thermoplastic but eliminate the ability to reshape. The engineering
thermosets include epoxies (for adhesives, structural composites, and
electronics encapsulation), phenolic resins (for laminates, brake
linings, and circuit-board substrates), polyester and vinyl-ester
resins (for fibre-reinforced composites), and polyurethanes (which
can be thermoset or thermoplastic depending on the formulation).

\engterm{Elastomers}. Lightly crosslinked networks with a glass
transition far below room temperature, so that the chains between
crosslinks are highly mobile. Stretching the elastomer extends the
chains; the entropy of the stretched configuration is lower than that
of the coiled configuration, and the chains pull back when the load
is released. The thermodynamic origin of rubber elasticity is entropic
rather than enthalpic, the reverse of the situation in metals and
ceramics. The engineering elastomers include natural rubber (cis-
polyisoprene), synthetic rubbers (styrene-butadiene rubber, nitrile
rubber, butyl rubber, chloroprene/neoprene), silicone elastomers
(polydimethylsiloxane), and fluoroelastomers (Viton family). The
crosslink density determines the modulus: typical engineering rubbers
have one crosslink per $10^2$ to $10^3$ monomer units, with elastic
modulus from $1\,\text{MPa}$ to $20\,\text{MPa}$.

\engterm{Crystallinity}. Polymers can be amorphous (no long-range
order, like glass) or semicrystalline (a partial crystalline fraction
in an amorphous matrix). The crystalline regions are typically
lamellae of folded chains; the amorphous regions are entangled
chains. The crystallinity fraction depends on chain regularity
(highly regular chains crystallise readily, irregular chains do not),
on cooling rate (fast cooling traps the amorphous state, slow cooling
allows crystallisation), and on annealing. Polyethylene's
crystallinity ranges from $\sim 40\,\%$ (LDPE, with extensive
branching) to $\sim 95\,\%$ (HDPE, linear); polypropylene is typically
$50$ to $70\,\%$ crystalline. Polystyrene and polymethyl methacrylate
are essentially fully amorphous in commercial grades because their
atactic structure prevents crystallisation.

\engterm{The chain entanglement network}. Above the molecular weight
$M_e$ (the entanglement molecular weight, typically $\sim 10^3$ to
$10^4\,\text{g}/\text{mol}$ for common polymers), chains are
constrained by their topological interactions with other chains. The
entanglement network behaves like a temporary crosslink: on short
time scales the chains are constrained, on long time scales they
slide and disentangle. The transition between the two regimes (the
disentanglement time $\tau_d$) is set by the chain length, the
temperature, and the chemistry; for polyethylene melts at processing
temperatures, $\tau_d$ is in the $0.1\,\si{\second}$ to $10\,
\si{\second}$ range. The entanglement network gives a polymer melt
its characteristic viscoelastic response and gives the solid polymer
its toughness above the brittle-ductile transition.

\section{Glass transition and viscoelasticity}\pathtag{core}

The most engineering-significant transition in a polymer is the
\engterm{glass transition} at temperature $T_g$. Below $T_g$ the
polymer is glassy: the chains are frozen in place, the material is
stiff and brittle, the modulus is high ($\sim 1$ to $5\,\text{GPa}$),
and the failure strain is small ($< 5\,\%$). Above $T_g$ the chains
have local mobility: the material is rubbery, the modulus drops by
two to three orders of magnitude, and the failure strain becomes
large (often $> 100\,\%$ in elastomers). The transition spans a
temperature range of $\sim 10$ to $20\,\si{\kelvin}$ rather than
occurring at a single temperature.

\engterm{Values of $T_g$ for engineering polymers}.

\begin{itemize}[leftmargin=*]
\item Polyethylene: $\sim -100$ to $-130\,\degC$ (deeply below room
  temperature, so room-temperature PE is in its rubbery region; the
  crystalline regions provide the stiffness).
\item Polypropylene: $\sim -10\,\degC$.
\item Polyvinyl chloride: $\sim 80\,\degC$ (rigid PVC at room
  temperature; the chlorine atoms increase intermolecular cohesion
  and raise $T_g$).
\item Polystyrene: $\sim 100\,\degC$.
\item Polymethyl methacrylate: $\sim 105\,\degC$.
\item Polycarbonate: $\sim 145\,\degC$.
\item Polyamide 6,6 (nylon 6,6): $\sim 50\,\degC$ in dry condition,
  lower when humid because absorbed water plasticises the polymer.
\item Polyethylene terephthalate (PET): $\sim 70\,\degC$.
\item Polyetheretherketone (PEEK): $\sim 145\,\degC$.
\item Natural rubber: $\sim -70\,\degC$.
\item Silicone rubber: $\sim -125\,\degC$.
\end{itemize}

\engterm{Kinetic vs thermodynamic origin}. The glass transition is a
kinetic transition rather than a thermodynamic one (no discontinuity
in the entropy or volume across the transition, only a discontinuity
in the slope of these quantities). The measured $T_g$ depends on the
cooling rate: faster cooling traps the polymer in a higher-temperature
configuration and produces higher $T_g$. The shift is typically a
few kelvins per decade of cooling rate change. The fictive temperature
(the temperature at which the glassy polymer would be in equilibrium
with its measured volume) is a useful conceptual tool but not
typically reported on data sheets.

\engterm{Viscoelasticity}. Above $T_g$, polymers exhibit a combination
of elastic and viscous response: when loaded, they deform partly
elastically (the deformation recovers when the load is removed) and
partly viscously (the deformation accumulates over time). The
simplest models are the Maxwell element (a spring in series with a
dashpot, used to describe stress relaxation) and the Voigt element
(a spring in parallel with a dashpot, used to describe creep). Real
polymers require a spectrum of relaxation times to describe their
behaviour; the spectrum is conventionally captured by the
\engterm{dynamic mechanical analysis} of the storage modulus $E'$ and
loss modulus $E''$ as functions of frequency and temperature.

\engterm{The time-temperature superposition principle}. The
viscoelastic response of a polymer at temperature $T$ and frequency
$\omega$ is equivalent to the response at a different temperature
$T_0$ and a shifted frequency $\omega a_T$, where $a_T$ is the
\engterm{shift factor}. For temperatures within $\sim 100\,\si{\kelvin}$
of $T_g$, the shift factor follows the Williams-Landel-Ferry equation,
\[
\log a_T = \frac{-C_1 (T - T_0)}{C_2 + (T - T_0)},
\]
with $C_1 \approx 17$ and $C_2 \approx 51\,\si{\kelvin}$ as universal
constants for many polymers when $T_0 = T_g$. The principle allows
the engineer to predict long-time creep behaviour from short-time
elevated-temperature measurements, or to extrapolate high-frequency
acoustic and dielectric measurements from quasi-static data.

\engterm{Creep and stress relaxation}. The two canonical viscoelastic
experiments. In a \engterm{creep test}, a constant stress is applied
and the strain is measured as a function of time; the creep
compliance $J(t) = \varepsilon(t)/\sigma_0$ increases with time and
reaches a steady-state slope (the steady-state creep rate) at long
times. In a \engterm{stress-relaxation test}, a constant strain is
applied and the stress is measured; the relaxation modulus $E(t) =
\sigma(t)/\varepsilon_0$ decreases with time and reaches a long-time
plateau (the rubbery plateau, set by the entanglement network). The
two tests are related but not identical; they can be transformed
into each other through the linear-viscoelastic constitutive
equation.

\section{Common engineering polymers}\pathtag{standard}

The chapter's working catalogue of polymers covers approximately
$95\,\%$ of plastic production by mass and the bulk of engineering
specifications.

\engterm{Polyethylene (PE)}. The most-produced polymer ($\sim 110\,
\text{Mt}/\text{yr}$ globally as of 2025). Repeat unit
$-\text{CH}_2-\text{CH}_2-$. Three major grades:

\begin{itemize}[leftmargin=*]
\item Low-density polyethylene (LDPE): branched, $\rho = 0.91$ to
  $0.94\,\text{g}/\text{cm}^3$, $\sigma_y \sim 10\,\text{MPa}$, high
  elongation. Used for plastic film, squeeze bottles, and packaging.
\item Linear low-density polyethylene (LLDPE): linear with controlled
  short branches, $\rho = 0.92$ to $0.94$, intermediate properties.
  Dominant grade for stretch wrap and grocery bags.
\item High-density polyethylene (HDPE): linear, $\rho = 0.94$ to
  $0.97$, $\sigma_y \sim 25\,\text{MPa}$, lower elongation than LDPE,
  higher stiffness. Used for milk jugs, pipes, fuel tanks.
\item Ultra-high-molecular-weight polyethylene (UHMWPE):
  $\bar{M}_w > 3 \times 10^6\,\text{g}/\text{mol}$, exceptional
  abrasion resistance and impact strength. Used for hip-joint
  bearings, ballistic armour, ski bottoms.
\end{itemize}

\engterm{Polypropylene (PP)}. Repeat unit
$-\text{CH}_2-\text{CH(CH}_3)-$. Production $\sim 80\,\text{Mt}/\text{yr}$.
$\rho = 0.90\,\text{g}/\text{cm}^3$ (the lightest commercial polymer),
$\sigma_y \sim 35\,\text{MPa}$, $T_m \sim 165\,\degC$. Three major
grades:

\begin{itemize}[leftmargin=*]
\item Homopolymer PP: stiff, low impact strength at room temperature
  and brittle below freezing. Used for rigid packaging, automotive
  interior trim, fibres, ropes.
\item Random copolymer PP (with $\sim 1$ to $7\,\%$ ethylene): lower
  melting point and clearer than homopolymer. Used for clear food
  containers.
\item Impact copolymer (heterophasic) PP: a blend of PP homopolymer
  with ethylene-propylene rubber phase. Used for automotive bumpers,
  battery cases, durable goods.
\end{itemize}

\engterm{Polystyrene (PS) and ABS}. Polystyrene repeat unit
$-\text{CH}_2-\text{CH(C}_6\text{H}_5)-$. General-purpose polystyrene
is clear, rigid, brittle. High-impact polystyrene (HIPS, polystyrene
with $5$ to $10\,\%$ polybutadiene rubber phase) is opaque, lower
stiffness, much higher impact resistance. ABS (acrylonitrile-butadiene-
styrene terpolymer) combines acrylonitrile for chemical resistance,
butadiene for impact resistance, and styrene for stiffness; ABS is
the dominant engineering thermoplastic for consumer products such as
Lego bricks, automotive interior components, household appliance
housings.

\engterm{Polycarbonate (PC)}. Linear polyester of bisphenol-A and
carbonate. $\rho = 1.20\,\text{g}/\text{cm}^3$, $\sigma_y \sim 65\,
\text{MPa}$, $T_g = 145\,\degC$, optically clear, exceptionally
high impact resistance (orders of magnitude greater than acrylic or
polystyrene). Used for safety eyewear, motorcycle helmet visors,
electronic enclosures, optical media (CDs and DVDs), structural
sheet for greenhouses and security glazing.

\engterm{Polyethylene terephthalate (PET)}. Polyester of ethylene
glycol and terephthalic acid. Used for beverage bottles ($\sim 75\,\%$
of PET production), polyester fibres for textiles, and engineering
thermoplastic grades reinforced with glass fibre. The crystallised
form is opaque and stiff (used for moulded parts); the amorphous form
is clear (used for bottles).

\engterm{Polyamide (nylon)}. Condensation polymers of diamines and
diacids (or of an amino acid). The most common grades are nylon 6,6
(from hexamethylenediamine and adipic acid) and nylon 6 (from
caprolactam). Hydrogen-bonded between chain amide groups, giving high
melting point ($\sim 260\,\degC$ for nylon 6,6) and high tensile
strength. Used for fibres (textiles, ropes, fishing line), engineering
mouldings (gears, bearings), films (food packaging).

\engterm{Polyetheretherketone (PEEK)}. Aromatic polymer with ether and
ketone linkages in the backbone. $T_g = 145\,\degC$, $T_m = 343\,\degC$,
$\sigma_y \sim 100\,\text{MPa}$, exceptional chemical resistance.
Used for aerospace structural parts, oil-and-gas downhole components,
biomedical implants. Cost $\sim \$80$ to $\$200/\text{kg}$ as of 2025
limits use to high-performance applications.

\engterm{Polytetrafluoroethylene (PTFE, Teflon)}. Fully fluorinated
ethylene polymer. $T_m = 327\,\degC$, exceptional chemical inertness,
very low coefficient of friction (the lowest of any solid against
itself, $\sim 0.05$), low surface energy. Used for non-stick coatings,
chemical-process gaskets and seals, electrical insulation in
high-frequency applications. PTFE cannot be melt-processed (the melt
viscosity is too high); the powder is compacted and sintered like a
ceramic.

\engterm{Polyvinyl chloride (PVC)}. Repeat unit $-\text{CH}_2-\text{CHCl}-$.
The most chlorinated commodity polymer. Rigid PVC (no plasticiser) is
used for plumbing pipe, window frames, exterior siding. Flexible PVC
(with $20$ to $40\,\%$ plasticiser, traditionally phthalate esters) is
used for cable insulation, hoses, flooring, medical tubing. The
chlorine content makes PVC flame-retardant; on decomposition, however,
it releases hydrogen chloride gas, which is corrosive and toxic.

\section{Polymer processing}\pathtag{standard}

A polymer part is made by shaping the polymer melt (for thermoplastics)
or by curing a liquid resin (for thermosets) into the desired
geometry. The dominant processing routes:

\engterm{Extrusion}. The polymer pellets are melted in a heated barrel
with a rotating screw and pushed through a shaping die. Continuous-
cross-section products are produced: pipes, tubing, profiles for
window frames, films, fibres. The extruder's screw provides melting
through viscous heating (mechanical work converted to heat by the
melt's viscosity), mixing through the screw's geometry, and pumping
through the pressure gradient along the screw. Extruders are sized by
screw diameter ($25$ to $300\,\text{mm}$ for industrial machines) and
length-to-diameter ratio ($20:1$ to $40:1$). Specialty extruders
(twin-screw, counter-rotating or co-rotating) provide better mixing
for compounding operations where additives and reinforcements are
incorporated.

\engterm{Injection moulding}. The polymer melt is forced into a closed
mould cavity, cooled to solidify, and ejected. Cycle times are
typically $5$ to $60\,\si{\second}$, with the bulk of the cycle being
cooling. The process produces complex three-dimensional parts in high
volume; injection-moulded parts dominate consumer electronics,
automotive interiors, medical disposables, and packaging. Injection-
moulding machine sizes range from a few grams of shot weight (for
medical devices and electronic connectors) to several kilograms (for
automotive bumpers and outdoor furniture). The major process variables
are melt temperature, mould temperature, injection pressure and rate,
cooling time; the major part-quality defects are sink marks (insufficient
hold pressure), short shots (cavity not filled), flash (mould flash
gap excessive), and weld lines (incomplete fusion at the meeting of
two flow fronts).

\engterm{Blow moulding}. A parison (a hollow tube of melt) is enclosed
in a mould cavity and inflated with air to take the cavity's shape.
The process is used for bottles, drums, fuel tanks, and other hollow
parts. Variants include extrusion blow moulding (the parison is
extruded continuously), injection blow moulding (the parison is
injection-moulded first), and stretch blow moulding (used for PET
beverage bottles, with biaxial orientation increasing the bottle's
strength and barrier properties).

\engterm{Thermoforming}. A sheet of thermoplastic is heated to its
softening range and shaped over a mould by vacuum or pressure. Used
for thin-walled containers, packaging trays, refrigerator and bathtub
liners. Lower tooling cost than injection moulding but limited to
relatively simple shapes.

\engterm{Compression and transfer moulding}. The thermoset resin (with
fillers and reinforcements) is loaded into a heated mould, the mould
is closed, and the resin cures under pressure. Used for thermoset
parts that cannot be processed by extrusion or injection moulding
(electrical connectors, brake linings, structural composites).

\engterm{Additive manufacturing (3D printing)}. A part is built layer
by layer from a digital geometry. The two dominant polymer additive
methods:

\begin{itemize}[leftmargin=*]
\item Fused filament fabrication (FFF), also called fused deposition
  modelling (FDM): a thermoplastic filament is melted in a moving
  extruder head and deposited along the layer's path. Used for
  prototyping, low-volume parts, and consumer-grade printing. Common
  materials include PLA (a bio-derived polyester), ABS, PETG, nylon,
  and PEEK.
\item Stereolithography (SLA) and digital light processing (DLP): a
  photopolymer resin is selectively cured by laser or projector
  exposure. Higher resolution than FFF, used for dental models,
  hearing-aid shells, jewellery casting masters.
\end{itemize}

Additive manufacturing has changed the economics of low-volume polymer
parts production, particularly for prototypes and for personalised
medical devices. The mechanical properties of additively manufactured
polymer parts depend strongly on the print parameters (layer height,
infill pattern, raster orientation) and are often anisotropic with the
weak direction perpendicular to the layer plane. Chapter~11 of this
volume develops the manufacturing-properties linkage further.

\section{Polymer degradation and ageing}\pathtag{standard}

Polymers degrade in service by chemical, thermal, mechanical, and
combined mechanisms. The degradation reduces molecular weight,
introduces chemical modifications, and changes mechanical, optical,
and electrical properties. Engineering polymer selection must
include the degradation pathway expected in the service environment
and the rate of property loss along that pathway.

\engterm{Oxidation}. Atmospheric oxygen reacts with polymer chains,
especially at carbon-hydrogen bonds adjacent to electron-rich groups.
The reaction proceeds by a free-radical chain mechanism: a hydrogen
atom is abstracted, a peroxide forms with atmospheric oxygen, the
peroxide decomposes to two new radicals, and the cycle propagates.
The chain reactions produce broken-backbone chain scission (lowering
molecular weight) and crosslinking (raising it), with the net effect
depending on the polymer. Polyethylene and polypropylene undergo
predominantly chain scission and become brittle; natural rubber
crosslinks and becomes stiff and rigid. The degradation is
accelerated by temperature (Arrhenius) and by transition-metal
impurities that catalyse peroxide decomposition.

\engterm{Photo-oxidation}. Ultraviolet radiation (wavelengths $290$ to
$400\,\text{nm}$ in sunlight) provides the activation energy for
oxidation reactions directly, bypassing the thermal pathway. Polymers
exposed to sunlight outdoors undergo rapid photo-oxidation. The
mechanism is similar to thermal oxidation but localised at the
exposed surface; the depth of penetration of UV radiation is typically
$10$ to $100\,\si{\micro\meter}$, so photo-degradation is a
surface-layer effect. The degraded surface layer becomes brittle,
cracks under mechanical or thermal cycling, and the cracks then
penetrate into the underlying bulk. Polymer formulations for outdoor
use include UV stabilisers (hindered amine light stabilisers, or
HALS, and benzotriazole UV absorbers) and antioxidants (hindered
phenols).

\engterm{Hydrolysis}. Polymers with hydrolysable backbone linkages
(esters in polyesters and polycarbonates, amides in nylons, urethanes
in polyurethanes) react with water, cleaving the backbone. The
reaction is slow at room temperature but accelerates above $T_g$ and
above $\sim 50\,\degC$. Polyester fibres in marine ropes lose strength
over years of submerged service; polycarbonate components in
high-humidity environments (boilers, autoclaves, tropical climates)
crack and yellow. The mitigation includes choosing more
hydrolysis-resistant chemistries (polyolefins instead of polyesters
for marine ropes), controlled humidity in service, and stabiliser
packages.

\engterm{Thermal degradation}. At elevated temperatures, polymer chains
undergo bond cleavage by thermolysis (no oxygen required). The
activation energy is high ($\sim 200$ to $300\,\text{kJ}/\text{mol}$),
so thermal degradation is slow at room temperature but rapid above
$\sim 300\,\degC$ for most polymers. Processing temperatures are
chosen to be high enough for melt flow but low enough to limit
thermal degradation during the residence time in the processing
machine; antioxidants are added to limit the degradation that does
occur.

\engterm{Radiation degradation}. Gamma rays, X-rays, electron beams,
and high-energy ultraviolet light can break polymer chains directly.
The effects are used deliberately (radiation crosslinking of
polyethylene to produce wire insulation with elevated-temperature
service capability, electron-beam curing of thin coatings) and avoided
in some applications (medical-device polymers sterilised by gamma
radiation must be formulated to retain properties after the dose).

\engterm{Plasticiser migration}. In plasticised polymers (notably
flexible PVC), the low-molecular-weight plasticiser diffuses out of
the polymer over time. The polymer becomes harder, more brittle, and
loses its flexibility. The migration is accelerated by elevated
temperature and by contact with absorbent surfaces. Plasticised PVC
medical tubing and food-packaging films are subject to plasticiser-
migration limits in regulatory regimes (FDA in the United States,
European Food Safety Authority in the European Union).

\engterm{Stress whitening and crazing}. A polymer under tensile stress
develops microscopic voids in the form of \engterm{crazes}: regions
of locally stretched fibrils bridging a gap perpendicular to the
applied stress. Crazes are precursors to cracks; they appear as
silver-coloured patches in transparent polymers (the colour comes
from light scattering off the fibrils) and as white opaque regions in
otherwise transparent or colourless polymer. Crazing is a
crack-tolerant deformation mode in glassy polymers; the work done in
creating the craze fibrils absorbs energy that would otherwise drive
crack propagation. Toughened polystyrene (HIPS, polystyrene with
butadiene rubber particles) and ABS exploit craze toughening as their
principal mechanism of impact resistance.

\section{Failure: creep, environmental stress cracking, photo-oxidation}\pathtag{core}

Three failure modes account for the bulk of polymer service failures
and each has a characteristic mechanism, time scale, and mitigation.

\engterm{Creep failure}. A polymer under sustained load deforms with
time, even at stresses below the short-term yield strength. The
deformation may be reversible (viscoelastic creep, recovering when
the load is removed) or irreversible (viscous flow, plastic
deformation, crack growth). At sufficient strain or sufficient time,
the polymer ruptures. The failure mode is dominant for:

\begin{itemize}[leftmargin=*]
\item Polyethylene pipe carrying water or natural gas under internal
  pressure. The pipe wall is in continuous hoop stress at
  approximately $5$ to $10\,\text{MPa}$, well below the polymer's
  short-term yield strength of $\sim 25\,\text{MPa}$. The service
  life is set by the slow creep rupture, with the standard ISO 9080
  procedure extrapolating short-term high-temperature creep-rupture
  data to the service temperature and time via the time-temperature
  superposition principle. Modern PE-100 pipe grades show
  $50$-year service life at $10\,\text{MPa}$ hoop stress.
\item Plastic structural components in outdoor furniture, agricultural
  equipment, and architectural fixtures. The sustained loads
  (gravity, wind, thermal expansion) accumulate creep strain over
  years, eventually producing visible deformation or failure.
\item Wire-tie cable management hardware (zip ties, cable clamps) in
  long-duration service. The tightened tie creates a sustained
  tensile load; the polymer creeps and eventually relaxes, allowing
  the bundle to loosen or the tie to break.
\end{itemize}

The mitigation includes specifying polymer grades with adequate
creep-rupture life at the service stress, designing for the
expected creep strain, and accepting periodic replacement for
components in long-duration high-stress service.

\engterm{Environmental stress cracking (ESC)}. A polymer under stress
in contact with a chemically aggressive environment cracks at
stresses far below the polymer's short-term yield strength. The
mechanism is a coupling between stress and chemistry: the stress
opens microscopic crazes or microcracks, the environment penetrates
into them, the chemical attack weakens the polymer locally, and the
crack propagates. ESC is the most common failure mode of
polyethylene in service and is the source of much of the historic
warranty experience with HDPE in service.

The standard ESC test, the Bell Telephone bent-strip test (ASTM
D1693), bends a polymer specimen over a former at a specified
strain and immerses it in a surface-active agent (Igepal CO-630, a
nonionic surfactant) at $50\,\degC$. The time to cracking is the
ESC resistance metric. Modern polyethylene grades for service in
contact with detergents, alcohols, or hydrocarbon-bearing
environments are formulated for extended ESC life: PE-100 grades show
ESC life exceeding $1000\,\text{h}$ in the bent-strip test, compared
to less than $100\,\text{h}$ for early high-density polyethylene.

\engterm{Photo-oxidation in outdoor service}. Polymers exposed to
direct sunlight degrade at rates depending on UV-A and UV-B
intensity, polymer chemistry, and stabiliser package. The visible
manifestations are colour change (yellowing of polycarbonate and
ABS, fading of pigmented PVC), surface chalking (loss of resin from
exposed surfaces, leaving pigment particles), and surface cracking
(craze networks that propagate inward over years). The failure
proceeds at the surface, with the bulk polymer unaffected until the
surface crack network produces structural compromise.

Polymer selection for outdoor service follows a hierarchy:

\begin{itemize}[leftmargin=*]
\item For maximum UV resistance, choose inherently stable chemistries:
  fluoropolymers (PVDF, ETFE for architectural cladding), acrylics
  (PMMA for outdoor signage and glazing), silicones (sealants and
  coatings).
\item For polyolefins (PE, PP) in outdoor service, add carbon black
  ($1$ to $3\,\%$ by mass) as a UV absorber; the black colour is
  the indicator of the formulation. Carbon-black-stabilised
  polyolefins show service lives of decades in outdoor service.
\item For visually clear or coloured outdoor applications, add HALS
  and UV absorber packages; expect service lives of $5$ to $15$
  years before recoating or replacement.
\end{itemize}

The accelerated weathering tests (xenon-arc and QUV cabinets per
ASTM G155 and G154) provide engineering correlations to outdoor
service, with the standard caveat that the laboratory acceleration
factor depends on the polymer-environment combination and on the
specific failure mode.

\begin{principle}[Polymer service life is a product of time and chemistry]
A polymer's service life in a chosen environment is the integral of
its degradation rate over the service condition's combined stress,
temperature, oxidant exposure, UV exposure, and hydrolytic exposure.
No single accelerated test substitutes for the full service-condition
prediction. The engineering response is to identify the dominant
degradation pathway, to choose a polymer formulation whose properties
along that pathway match the required service life, and to specify
end-of-life criteria that allow inspection-based replacement before
catastrophic failure.
\end{principle}

\begin{estimation}
\textbf{Question.} Estimate the difference in flexural modulus between
polystyrene at room temperature ($25\,\degC$) and at $80\,\degC$.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
Polystyrene has $T_g \approx 100\,\degC$. At $25\,\degC$ the polymer
is glassy and stiff: handbook value of flexural modulus is $\sim 3\,
\text{GPa}$. At $80\,\degC$ the polymer is still glassy (below $T_g$)
but approaching the glass transition; the modulus is reduced but not
collapsed.

A first estimate uses the standard time-temperature superposition rule:
the modulus drops by a factor of $\sim 100$ to $1000$ across the
glass transition, which spans $\sim 20\,\si{\kelvin}$. At $25\,\degC$
we are $75\,\si{\kelvin}$ below $T_g$, in the deep glassy region; at
$80\,\degC$ we are $20\,\si{\kelvin}$ below $T_g$, at the upper edge
of the glassy region. The modulus might drop by a factor of $1.5$ to
$3$ across this range, from $3\,\text{GPa}$ to $\sim 1$ to $2\,
\text{GPa}$.

Measured handbook data: polystyrene flexural modulus is approximately
$3.2\,\text{GPa}$ at $25\,\degC$ and $\sim 2.0\,\text{GPa}$ at
$80\,\degC$. The estimate captures the magnitude of the reduction.

\smallskip
\textit{Postmortem.} The estimate is internally consistent with the
qualitative picture: glassy polymers retain most of their stiffness up
to within $\sim 30\,\si{\kelvin}$ of $T_g$ and then lose stiffness
abruptly. The chapter's broader engineering claim is that polymer
service temperatures must be at least $\sim 30$ to $50\,\si{\kelvin}$
below $T_g$ for the polymer's stiffness to be useful; the closer one
gets to $T_g$, the more important creep and stress-relaxation
behaviour become. The estimate gives the working order of magnitude
without requiring a viscoelastic model.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Characterise the stress-strain curve of three polymers]
The reader selects three household polymer objects (or samples of
polymer-rich material), pulls each to failure in a simple tensile
test, and constructs the stress-strain curve for comparison.

\textbf{Track}. Physical characterisation, household instrumentation.
\textbf{Hazard class}: low. Hand-pulling of polymer strips; safety
glasses recommended for the failure event.

\textbf{Samples}. Select three with substantially different polymer
classes:

\begin{itemize}[leftmargin=*]
\item A polyethylene grocery bag (LDPE or LLDPE).
\item A polyester drinking-bottle wall (PET).
\item A rubber band (natural or synthetic elastomer).
\item Optional: a polystyrene drinking-cup wall (PS), a polypropylene
  food-container wall (PP), a nylon fishing line.
\end{itemize}

\textbf{Procedure}.

\begin{enumerate}[leftmargin=*]
\item Cut each sample to a uniform strip ($\sim 10\,\text{mm}$ wide,
  $\sim 100\,\text{mm}$ long, with parallel sides) using a sharp
  blade and a straight edge.
\item Measure the cross-sectional dimensions (width and thickness) at
  three points along the length and compute the average area.
\item Mount the strip in a simple tensile fixture: a kitchen scale or
  spring scale capable of reading at least $200\,\text{N}$, clamped
  to one end of the strip; the other end held by hand or clamped to
  a fixed support. A digital fish scale works.
\item Mark two gauge lines $50\,\text{mm}$ apart on the strip.
\item Apply tensile force slowly while reading the force from the
  scale and the gauge length with a ruler. Record force-elongation
  pairs at $\sim 10$ to $20$ points up to failure.
\item Compute engineering stress $\sigma = F/A_0$ and engineering
  strain $\varepsilon = \Delta L/L_0$.
\item Plot stress vs strain for each sample.
\end{enumerate}

\textbf{Analysis}.

\begin{itemize}[leftmargin=*]
\item For each sample, identify the elastic region (linear initial
  portion), the yield point (where the linear region ends), the
  ultimate stress (peak), and the strain at failure.
\item Compute the elastic modulus from the initial slope.
\item Compare the three curves qualitatively: which is most ductile,
  which is most stiff, which is most extensible?
\item Identify which polymer the sample most likely is, based on its
  stress-strain signature.
\item For each sample, identify which polymer class (thermoplastic,
  thermoset, elastomer) the curve is consistent with and which
  chain or network feature explains the qualitative shape.
\end{itemize}

\textbf{Deliverable}. A four- to seven-page report with the three
stress-strain curves overlaid on a single plot, the property table
(modulus, yield stress, ultimate stress, strain at failure), and a
paragraph per sample tying the curve shape to the polymer's chain
or network structure.

\textbf{Time}. Two to three hours of bench work, two hours of
write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A polyethylene sample has number-average molecular weight $\bar{M}_n
= 50\,000\,\text{g}/\text{mol}$ and weight-average molecular weight
$\bar{M}_w = 150\,000\,\text{g}/\text{mol}$. Compute the polydispersity
index and the average degree of polymerisation (use ethylene monomer
mass $28\,\text{g}/\text{mol}$).
\end{exercise}

\begin{exercise}
A polypropylene fibre has tensile strength $600\,\text{MPa}$. Compute
the force required to break a fibre with cross-section $0.05\,\text{mm}^2$.
Compare to a steel wire of the same cross-section.
\end{exercise}

\begin{exercise}
The Williams-Landel-Ferry shift factor at $T = T_g + 20\,\si{\kelvin}$
is computed with $C_1 = 17.4$ and $C_2 = 51.6\,\si{\kelvin}$. Compute
$\log_{10} a_T$ and interpret in terms of the time-temperature
superposition for a creep test.
\end{exercise}

\begin{exercise}
A lightly crosslinked rubber has crosslink density $\nu = 10^{-4}\,
\text{mol}/\text{cm}^3$. Estimate the shear modulus from the
rubber-elasticity relation $G = \nu R T$ at room temperature.
\end{exercise}

\begin{exercise}
A polyethylene pipe with $50\,\text{mm}$ outer diameter and $5\,\text{mm}$
wall thickness operates at $8\,\text{bar}$ internal pressure. Compute
the hoop stress and compare to the polymer's $50$-year creep-rupture
stress limit of $8\,\text{MPa}$ at $20\,\degC$.
\end{exercise}

\begin{exercise}
A polycarbonate safety eyewear lens absorbs $0.2\,\text{J}$ of impact
energy without fracture. Estimate the strain at the absorbed-energy
limit using a yield stress of $65\,\text{MPa}$ and a Young's modulus
of $2.3\,\text{GPa}$, and a lens volume of $5\,\text{cm}^3$.
\end{exercise}

\begin{exercise}
An injection-moulded ABS housing weighs $250\,\text{g}$ and has wall
thickness $2.5\,\text{mm}$. Compute the cooling time required to
solidify the part from melt temperature ($240\,\degC$) to ejection
temperature ($80\,\degC$), given thermal diffusivity $\alpha = 1.2
\times 10^{-7}\,\text{m}^2/\si{\second}$. Use the scaling $t \sim
L^2/\alpha$.
\end{exercise}

\begin{exercise}
A natural-rubber band stretched to $\lambda = 4$ times its original
length develops a tension of $1.5\,\text{N}$. Estimate the rubber's
shear modulus from the neo-Hookean expression $\sigma = G(\lambda^2
- 1/\lambda)$ with the rubber's cross-section $1\,\text{mm}^2$.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the rubber-elasticity relation $G = \nu R T$ for an ideal
elastomer with crosslink density $\nu$, starting from the entropy of
a Gaussian chain.
\end{exercise}

\begin{exercise}
Show that a Maxwell element (spring in series with dashpot) produces
an exponential stress relaxation $\sigma(t) = \sigma_0 \exp(-t/\tau)$
under a step strain, with $\tau$ the spring constant divided by the
dashpot's viscosity.
\end{exercise}

\begin{exercise}
Derive the Williams-Landel-Ferry equation $\log a_T = -C_1(T - T_0)/
(C_2 + T - T_0)$ from a free-volume model of segmental relaxation.
\end{exercise}

\begin{exercise}
Show that the time-temperature superposition principle implies that
a polymer's creep compliance $J(T_1, t)$ at temperature $T_1$ and time
$t$ equals $J(T_0, t/a_T)$ at the reference temperature $T_0$ and
shifted time $t/a_T$.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the number of CH$_2$ repeat units in a polyethylene chain of
molecular weight $300\,000\,\text{g}/\text{mol}$, and estimate the
fully extended chain length using bond length $1.54\,\si{\angstrom}$
and bond angle $109.5^{\circ}$.
\end{exercise}

\begin{exercise}
Estimate the time required for a $30\,\%$ creep strain to develop in
a polyethylene pipe at $20\,\degC$ under $5\,\text{MPa}$ hoop stress,
given that creep at $80\,\degC$ produces the same strain in
$100\,\text{h}$ and the WLF shift factor between $80\,\degC$ and
$20\,\degC$ is $\sim 10^4$.
\end{exercise}

\begin{exercise}
Estimate the photo-oxidation depth penetration in polyethylene
exposed to noon sunlight, given the UV-B absorption coefficient
$\alpha \approx 50\,\text{cm}^{-1}$ at $310\,\text{nm}$.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a Rouse chain model for polymer dynamics: $N$ beads
connected by Hookean springs, undergoing Brownian motion in a
viscous solvent. Compute the mean-square end-to-end distance versus
time and identify the longest relaxation time $\tau_R$.
\end{exercise}

\begin{exercise}
Implement a viscoelastic finite-element model of a polymer
component (a simple beam or cantilever). Apply a step stress and
compute the time evolution of the strain. Verify against a
creep-compliance handbook value.
\end{exercise}

\begin{exercise}
Implement a chain-growth polymerisation Monte Carlo simulation.
Initiate $100$ chains and propagate to a chosen conversion. Plot the
molecular-weight distribution and compute $\bar{M}_n$ and $\bar{M}_w$.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify the polymer and the wall thickness for a $1\,\text{L}$
beverage bottle that must survive a $1\,\text{m}$ drop test fully
loaded, contain carbonated content at $4\,\text{bar}$ for
$6\,\text{months}$, and be transparent. Identify the candidate
chemistries and the dominant property requirement for each.
\end{exercise}

\begin{exercise}
Specify the polymer for a $50\,\text{m}$-long buried gas-distribution
pipe with $63\,\text{mm}$ diameter, $5.7\,\text{mm}$ wall, $4\,\text{bar}$
operating pressure, and $50$-year service life at $20\,\degC$. State
the standard (typically ISO 4437 or PE 100 specifications) and the
welding method.
\end{exercise}

\begin{exercise}
Specify the polymer for an automotive instrument-panel skin that
must survive $10$ years of outdoor sunlight exposure through the
windshield, with daily thermal cycling between $-30\,\degC$ and
$+85\,\degC$ and resistance to scratch from contact with passengers'
clothes. Identify the additive package.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A plastic outdoor chair shows fine craze networks on its upper
surfaces after one summer of outdoor use. Identify the polymer
(consistent with the visual evidence) and the most likely failure
mechanism. Propose two formulation changes that would extend service
life.
\end{exercise}

\begin{exercise}
A nylon gear in a power tool fails after six months of intermittent
use. The failure surface shows ductile tearing rather than brittle
fracture. Identify the polymer-specific failure mode and propose a
material change.
\end{exercise}

\begin{exercise}
A polyethylene grocery bag has tensile strength $25\,\text{MPa}$ when
tested fresh and $5\,\text{MPa}$ after one month of outdoor exposure.
Identify the most likely degradation pathway and propose a single
formulation change that would have prevented the loss.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A PVC plumbing pipe at a hot-water service location fails by brittle
cracking along the longitudinal direction after seven years of
service. The same brand and grade of pipe is in service in
cold-water applications without failures. Identify the failure mode
specific to elevated-temperature service and propose a service-
temperature limit.
\end{exercise}

\begin{exercise}
A polycarbonate safety helmet fails by brittle fracture during a
moderate-impact event. The helmet had been in service for $10$ years
in outdoor industrial use. Identify the time-dependent failure
mechanism that lowered the impact resistance below the rated
specification.
\end{exercise}

\begin{exercise}
An ABS automotive grille becomes chalky and pale after five years of
outdoor service. The polymer matrix is intact but the surface gloss
has been lost. Identify the surface-degradation mechanism and
propose a cosmetic refurbishment vs replacement decision based on
the underlying integrity of the structural component.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A colleague proposes replacing a $4140$ steel shaft with a glass-
fibre-reinforced nylon shaft of equal diameter to save mass.
Identify the property differences (stiffness, fatigue behaviour,
thermal expansion, creep at elevated temperature) that the
substitution would introduce and the application factors that
would make the substitution acceptable or not.
\end{exercise}

\begin{exercise}
A vendor claims a new biodegradable polymer has yield strength and
modulus comparable to nylon and biodegrades fully in eighteen
months in soil. Identify the qualifications the vendor would need
to document (mechanical properties across temperature, humidity
behaviour, degradation rate in non-soil environments) and the
risks of the qualification being achieved only in the laboratory
conditions of testing.
\end{exercise}

\begin{exercise}
A student argues that polymer creep is ``just like'' metal creep
and follows the same Arrhenius rate law. Identify the
qualifications this claim requires (polymer creep has a glass-
transition-dominated mechanism distinct from the dislocation-climb
mechanism of metal creep) and the working range over which the
analogy is approximately valid.
\end{exercise}
