\chapter{Entropy and the arrow of time}
\label{vol04:ch04}

\epigraph{If we wish to find in rational mechanics an a priori
foundation for the principles of thermodynamics, we must seek
mechanical definitions of temperature and entropy.}{Ludwig Boltzmann,
\emph{Lectures on Gas Theory} (1896), in the Brush translation
\cite[p.~5]{hist:boltzmann1896}.}

\chapmeta{Half-life: foundational. Archetypes: stability, uncertainty. Exercise target: 26. Project track: stochastic simulation of an ideal gas. Hazard class: none. Reader path default: standard, with sections explicitly tagged. Named cases: none required in this chapter; statistical-fluctuation and systematic-violation patterns are described by mechanism.}

\noindent
Entropy in chapter~1 was a bookkeeping quantity: a state function
$S$ whose change along a reversible heat-transfer path is
$\delta Q/T$, and whose generation in any real process is non-
negative. The present chapter gives entropy a microscopic meaning,
extends the second law to its statistical interpretation, sketches
the free-energy machinery for chemical and biological equilibrium,
and closes by examining the boundary between a thermodynamic
violation (impossible) and a statistical fluctuation (rare).

\section{Microscopic interpretation: Boltzmann}\pathtag{core}

Ludwig Boltzmann's 1877 identification of entropy with the logarithm
of the number of microscopic states accessible to a system at fixed
macroscopic conditions is the foundation of statistical mechanics
and the source of all subsequent microscopic interpretations of
thermodynamic quantities \cite{hist:boltzmann1896}. For a system at
fixed energy, volume, and particle number with $W$ accessible
microstates,
\[
S = k_B \ln W,
\]
where $k_B = 1.381 \times 10^{-23}\,\text{J}/\si{\kelvin}$ is the
Boltzmann constant. The formula appears on Boltzmann's grave in
Vienna. The dimensions on both sides match: entropy in $\text{J}/
\si{\kelvin}$ on the left, with $k_B$ converting the dimensionless
logarithm of a count to the same units.

The microstate count $W$ is enormous for any macroscopic system.
One mole of an ideal gas at standard conditions has $W$ of order
$10^{10^{25}}$, so $\ln W$ is itself of order $10^{25}$ and $S$ is of
order $k_B \times 10^{25} \sim 10^2\,\text{J}/\si{\kelvin}$. The
absolute size of $S$ for ordinary systems is unimportant; the change
in $S$ between two macroscopic states is the engineering quantity.

The second law in microscopic form reads: an isolated system
spontaneously evolves toward macrostates that have more accessible
microstates. The reason is statistical: at any instant, the system
is in some specific microstate, and the trajectory through phase
space, governed by the deterministic equations of motion, samples
microstates roughly uniformly subject to conservation laws. A
macrostate with $10^{20}$ times more microstates is overwhelmingly
more likely to be visited than its less-populated alternative, and
once visited, the system spends nearly all its time in it.

\engterm{Counting microstates: combinatorics}. The simplest non-
trivial example: $N$ distinguishable particles distributed between
two equal boxes. The number of microstates with exactly $n$
particles in the left box is $W_n = \binom{N}{n}$. The most-populated
macrostate is the one with $n \approx N/2$, where the binomial
coefficient peaks; the entropy of this macrostate is $S = k_B \ln
\binom{N}{N/2} \approx k_B N \ln 2$ for large $N$. Any macrostate
with $n$ significantly different from $N/2$ has exponentially fewer
microstates and exponentially lower entropy. For $N = 10^{20}$, a
configuration with $n = N/2$ is more probable than $n = (N/2)(1 +
10^{-9})$ by a factor that is so large it is not worth writing down.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-microstate-counting.tex}

The mixing-entropy result of chapter~1 follows immediately. Removing
the partition between two gases at the same temperature and pressure
allows the system to access a much larger number of microstates: each
of the $N_A + N_B$ particles can now occupy either half of the
combined volume, where before each was restricted to its original
half. The increase in $\ln W$ is the mixing entropy. Figure~\ref{fig:vol04ch04:entropy-of-mixing}
plots the molar mixing entropy of a binary ideal mixture against
composition; it is zero at the pure endpoints and maximal at the
equimolar point, the same combinatorial peak that
figure~\ref{fig:vol04ch04:microstate-counting} shows for the
two-box count.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-entropy-of-mixing.tex}

\engterm{The Gibbs paradox and indistinguishability}. The classical
counting argument has a problem: if the two gases being mixed are
identical, the mixing entropy is calculated to be positive when (by
state-function arguments) it must be zero. The resolution requires
treating identical particles as indistinguishable; the corrected
count divides by $N!$ for any permutation of $N$ identical particles
that yields the same microstate. With the Gibbs factor $1/N!$
included, the calculated mixing entropy is zero for identical gases
and positive for distinguishable ones, as physics demands. The
resolution is one of the historical pointers to quantum statistics:
identical-particle indistinguishability is a quantum statement that
classical counting did not capture.

\begin{principle}[The second law is a statistical statement]
The second law of thermodynamics is not a mechanical law to be
derived from the equations of motion. It is a statistical statement
about the relative abundance of microstates corresponding to
different macrostates. The law's apparent absoluteness is the
practical truth that, for macroscopic $N$, the more populated
macrostate is more abundant by exponentially many factors, and any
spontaneous fluctuation toward the less-populated macrostate is so
unlikely that it would not be expected within the age of the
universe. Macroscopic time-asymmetry has its origin in this
combinatorial asymmetry plus the initial conditions of the system.
\end{principle}

\engterm{The Clausius inequality, derived}. Chapter~1 took the
entropy state function and its non-negative generation as given. The
macroscopic ground under that assumption is the Clausius inequality,
$\oint \delta Q/T \le 0$ for any cyclic process, and it follows from
the Kelvin-Planck statement alone by a construction worth carrying in
full, because the same construction reappears whenever a second-law
bound must be proved rather than asserted.
Figure~\ref{fig:vol04ch04:clausius-inequality} shows the apparatus.
Consider an arbitrary device executing a cycle while it exchanges heat
elements $\delta Q_k$ with reservoirs at temperatures $T_k$. We do not
supply those heat elements directly. Instead each $\delta Q_k$ is
delivered to the device by a reversible Carnot bridge that draws heat
from one common reference reservoir held at $T_0$ and rejects it to the
device at $T_k$. A reversible bridge delivering $\delta Q_k$ at $T_k$
must, by the Carnot relation $\delta Q_0/T_0 = \delta Q_k/T_k$, draw
\[
\delta Q_0 = T_0\,\frac{\delta Q_k}{T_k}
\]
from the reference reservoir.

Run the composite, the device plus all its bridges, through one
complete cycle. The device returns to its starting state, and every
bridge returns to its starting state, so the composite has zero net
internal-energy change. The only heat crossing the composite's outer
boundary is the total drawn from the single reservoir $T_0$, namely
$Q_0 = \sum_k \delta Q_0 = T_0 \sum_k \delta Q_k/T_k = T_0 \oint
\delta Q/T$ in the limit of many small elements. By the first law the
composite's net work output equals this heat,
\[
W_{\text{net}} = Q_0 = T_0 \oint \frac{\delta Q}{T}.
\]
The composite now exchanges heat with one reservoir only and produces
$W_{\text{net}}$ in a cycle. The Kelvin-Planck statement forbids a
cyclic device from delivering net positive work while exchanging heat
with a single reservoir, so $W_{\text{net}} \le 0$, and since $T_0 >
0$,
\[
\oint \frac{\delta Q}{T} \le 0,
\]
the Clausius inequality. Equality holds when every step of the
original device is itself reversible, because then the composite could
be run backward to give $W_{\text{net}} \ge 0$ as well, forcing
$W_{\text{net}} = 0$. For a reversible cycle, $\oint \delta Q_{\text{rev}}/T = 0$,
which makes $\delta Q_{\text{rev}}/T$ an exact differential and defines
the entropy state function $dS = \delta Q_{\text{rev}}/T$ that
chapter~1 used without proof. For an irreversible cycle the strict
inequality holds, and the gap from zero is exactly the entropy the
real cycle generates: writing any real path from state $a$ to state
$b$ and returning along a reversible path gives $\int_a^b \delta Q/T
\le S_b - S_a$, the working inequality that underwrites every entropy
balance in this volume. The microscopic counting of section~4.1 and
the macroscopic Clausius inequality are the two faces of the second
law: the count says why the inequality holds, and the inequality says
how to use it in a heat-and-work ledger.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-clausius-inequality.tex}

The cleanest demonstration of the combinatorial origin of entropy is
the free (Joule) expansion of figure~\ref{fig:vol04ch04:free-expansion}.
A gas held in one half of an insulated rigid container is released into
the whole container. No heat crosses the boundary and no work is done
on the surroundings, so for an ideal gas the temperature does not
change. Yet the entropy rises by $\Delta S = N k_B \ln 2$, because each
molecule now reaches twice the volume and the microstate count
multiplies by $2^N$. The free expansion isolates the statistical
content of the second law: entropy increases with neither heat nor work,
driven purely by the gain in accessible microstates.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-free-expansion.tex}

\begin{archetype}[Stability under uncertainty: the macrostate that survives]
The equilibrium macrostate is the one that survives sampling. A system
left alone visits microstates, and the macrostate with the most
microstates is the one it spends nearly all its time in. The width of
the entropy peak sets the size of the fluctuations: for $N$ particles
the relative fluctuation in any extensive quantity scales as
$N^{-1/2}$, so a macroscopic system is stable to the relative precision
$10^{-12}$ while a hundred-particle system fluctuates by a few percent.
This is the recurring engineering pattern behind every equilibrium
argument in the volume: the stable state is the high-multiplicity state,
its stability grows with system size, and the residual uncertainty is
quantified by the curvature of the entropy at its maximum. The same
pattern returns wherever a design relies on a large ensemble averaging
out individual randomness: thermal noise floors, statistical tolerancing,
and the law of large numbers in reliability all read off the
$N^{-1/2}$ scaling.
\end{archetype}

\section{Statistical mechanics for engineers}\pathtag{standard}

The statistical-mechanics machinery that engineering practice
demands has three working components: the canonical-ensemble
partition function, the Boltzmann distribution, and the
thermodynamic relations between partition function and
thermodynamic state functions.

\engterm{The canonical ensemble}. A system in thermal contact with a
reservoir at temperature $T$ visits microstates with probability
proportional to the Boltzmann factor $\exp(-E_i/k_B T)$, where $E_i$
is the energy of microstate $i$. The proportionality constant is
fixed by normalising the total probability to one, giving the
\engterm{Boltzmann distribution}
\[
P_i = \frac{e^{-E_i/k_B T}}{Z}, \quad Z = \sum_i e^{-E_i/k_B T}.
\]
The sum $Z$ is the \engterm{partition function}, the central
calculation object of statistical mechanics. For a system with
continuous energy levels, the sum becomes an integral over phase
space with appropriate measure.

The Boltzmann factor in velocity space is the Maxwell-Boltzmann speed
distribution of figure~\ref{fig:vol04ch04:maxwell-boltzmann}. Setting
$E = \tfrac12 m v^2$ and weighting by the $v^2$ density of states in
three dimensions gives $f(v) \propto v^2 e^{-mv^2/2k_B T}$, whose peak
shifts upward and whose width grows as temperature rises. The broadening
is the microscopic content of heat capacity: more thermal energy is
distributed across a wider band of molecular speeds.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-maxwell-boltzmann.tex}

\engterm{Worked example: the two-level system}. The simplest partition
function with a closed form has two microstates, a ground state at
energy $0$ and an excited state at energy $\epsilon$. The partition
function is $Z = 1 + e^{-\epsilon/k_B T}$, the excited-state probability
is $p_1 = e^{-\epsilon/k_B T}/Z = 1/(1 + e^{\epsilon/k_B T})$, the
average energy is $\langle E \rangle = \epsilon p_1$, and the heat
capacity is
\[
\frac{C}{k_B} = \left(\frac{\epsilon}{k_B T}\right)^2
\frac{e^{\epsilon/k_B T}}{\left(1 + e^{\epsilon/k_B T}\right)^2}.
\]
The heat capacity vanishes at both low temperature (the gap is too
large to excite) and high temperature (both states are equally
populated and no further heat can be stored), peaking near
$k_B T \approx 0.42\,\epsilon$. This is the Schottky anomaly of
figure~\ref{fig:vol04ch04:two-level-system}. The code in
\texttt{code/partition\_function.py} computes the full canonical
thermodynamics for any discrete spectrum and locates the Schottky peak
by bisection; the dataset \texttt{data/two\_level\_thermo.csv} is its
temperature sweep. The two-level model is the working skeleton of
paramagnets, defect populations, and spin systems.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-two-level-system.tex}

The partition function generates thermodynamic state functions. For
the canonical ensemble,
\begin{align*}
\langle E \rangle &= -\frac{\partial \ln Z}{\partial \beta}, \quad
\beta = \frac{1}{k_B T}, \\
F &= -k_B T \ln Z, \\
S &= -\left(\pdv{F}{T}\right)_V = k_B \ln Z + \frac{\langle E
\rangle}{T},
\end{align*}
where $F$ is the Helmholtz free energy of chapter~1. The
calculation of any thermodynamic state function reduces to the
calculation of $Z$ for the system's microscopic Hamiltonian.

\engterm{The partition function for an ideal gas}. For $N$ non-
interacting molecules in a volume $V$ at temperature $T$, with each
molecule having translational, rotational, and vibrational degrees of
freedom, the partition function factorises:
\[
Z = \frac{1}{N!} \left(\frac{V}{\Lambda^3}\right)^N z_{\text{int}}^N,
\]
where $\Lambda = h/\sqrt{2\pi m k_B T}$ is the thermal de Broglie
wavelength and $z_{\text{int}}$ accounts for the internal modes.
The $N!$ corrects for indistinguishability. The result, used with
the relations above, recovers the ideal-gas equation of state $pV =
N k_B T$ and the temperature-dependence of $c_v$ from the internal
modes.

\engterm{Equipartition}. For a system in thermal equilibrium at
temperature $T$, each quadratic degree of freedom in the
Hamiltonian contributes $\frac{1}{2} k_B T$ to the average energy and
$\frac{1}{2} k_B$ to the heat capacity at constant volume. For a
monatomic gas (three translational degrees per molecule), $\langle E
\rangle = \frac{3}{2} N k_B T$ and $C_v = \frac{3}{2} N k_B$. For a
diatomic gas at moderate temperature (three translational and two
rotational, with vibrational frozen out), $C_v = \frac{5}{2} N k_B$.
The equipartition theorem fails at low temperatures where quantum
effects freeze out individual modes (the heat capacity of hydrogen
gas drops at low temperature because the rotational modes freeze
out below a characteristic temperature of $\sim 85\,\si{\kelvin}$).

The statistical-mechanics machinery is the bridge between molecular
properties (masses, bond strengths, intermolecular potentials) and
bulk thermodynamic data (heat capacities, equations of state,
phase-transition temperatures). For engineering practice at
ordinary temperatures and densities, the bridge is one-way: the
tables in chapter~2 provide the bulk data directly, and statistical
mechanics provides the underwriting reason that those tables can be
constructed. For engineering practice in unusual regimes (cryogenic
fluids, plasmas, semiconductors, biological macromolecules), the
statistical-mechanics calculation is the route by which the tables
themselves are computed.

\engterm{Fluctuations and thermal noise}. Equipartition assigns each
quadratic degree of freedom an average energy $\tfrac12 k_B T$, and the
canonical ensemble says more: that average is the centre of a distribution
whose width the same $k_B T$ sets. A degree of freedom does not sit at its
mean energy; it fluctuates around it, and the size of the fluctuation is a
measurable quantity that surfaces as noise in every instrument sensitive
enough to see it. The general result reads the variance of the energy off
the curvature of the partition function, $\langle (\Delta E)^2\rangle =
k_B T^2 C_V$, tying the mean-square energy fluctuation to the heat capacity:
a system that stores heat readily fluctuates widely, and a system with small
heat capacity holds its energy tightly. The relative fluctuation
$\sqrt{\langle(\Delta E)^2\rangle}/\langle E\rangle$ scales as $N^{-1/2}$,
the same scaling the stability archetype of section~4.1 drew, so a
macroscopic body's energy is steady to a part in $10^{12}$ while a
hundred-molecule cluster wobbles by a percent.

The electrical instance is the one an engineer meets first. A resistor at
temperature $T$ generates a fluctuating voltage across its terminals with no
applied signal, the Johnson-Nyquist noise, whose mean-square value in a
bandwidth $B$ is
\[
\langle v_n^2\rangle = 4 k_B T R B,
\]
with $R$ the resistance. The noise is the equilibrium thermal motion of the
resistor's charge carriers read out as a voltage, and its power delivered to
a matched load, $k_B T B$, carries the temperature directly and depends on
neither the resistance nor the material. Figure~\ref{fig:vol04ch04:thermal-noise}
plots the root-mean-square noise voltage against resistance at room
temperature and at liquid-nitrogen temperature, the square-root law that sets
the floor beneath every low-level measurement. A voltmeter reading a megohm
source across a megahertz of bandwidth sees microvolts of noise it can never
remove, because the noise is the resistor obeying the same equipartition that
fixes its heat capacity. Cooling the front end drops the floor as
$\sqrt{T}$, which is why the most sensitive amplifiers, radio-astronomy
receivers, and quantum-device readouts run cold. The thermal-noise floor is
the second law asserting itself inside a circuit: the fluctuations cannot be
suppressed without lowering the temperature, and lowering the temperature
costs the refrigeration work the earlier case studies priced.

\begin{mastery}
\textbf{The fluctuation-dissipation theorem}. Johnson-Nyquist noise is one
face of a general identity that binds the random fluctuations of a system in
equilibrium to the way that system dissipates energy when driven. The
fluctuation-dissipation theorem states that the spectrum of a quantity's
spontaneous equilibrium fluctuations is fixed by the same response function
that governs how the quantity relaxes after a small push, weighted by $k_B T$.
For the resistor the dissipative response is the resistance $R$, the real
part of the impedance that turns current into heat, and the theorem returns
the noise spectral density $4 k_B T R$ from it; a component that dissipates
must fluctuate, and a component that does not dissipate cannot generate
thermal noise. A capacitor, purely reactive, has no thermal voltage noise of
its own for exactly this reason. The theorem is the reason a mechanical
oscillator's Brownian jitter is set by the same damping coefficient that
drains its ringing, why a galvanometer's needle trembles in proportion to its
own friction, and why the drag on a colloidal particle and its diffusion
constant are locked together by the Einstein relation $D = k_B T/\gamma$, the
original fluctuation-dissipation statement. The engineering lesson is that
noise and loss are not independent design targets: any mechanism added to a
system to dissipate unwanted energy also injects thermal fluctuation at the
rate $k_B T$ per mode, and the only way to lower the fluctuation floor of a
dissipative element is to lower its temperature. The theorem is Onsager
reciprocity's near neighbour, both descended from the time-reversal symmetry
of the microscopic dynamics, but it speaks about the size of fluctuations
where reciprocity speaks about the symmetry of cross-couplings.
\end{mastery}

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-thermal-noise.tex}

\begin{mastery}
\textbf{Negative absolute temperature}. A system whose energy is
bounded above can be driven into a state in which the higher-energy
microstates are more populated than the lower-energy ones. For the
two-level system, the ratio $p_1/p_0 = e^{-\epsilon/k_B T}$ exceeds one
only when $T$ is negative. A negative absolute temperature is hotter
than any positive temperature, not colder than absolute zero: heat
flows from a negative-temperature system to any positive-temperature
system placed in contact with it. The thermodynamic temperature is
defined through $1/T = (\partial S/\partial E)_{V,N}$, and for a
bounded-energy spectrum the entropy decreases as energy increases past
the half-filled point, making the derivative negative. Population
inversion in lasers is the engineering instance: the gain medium is
prepared with more atoms in the upper state than the lower, a
negative-temperature condition for the relevant levels, and stimulated
emission extracts work from the inverted population. Negative
temperature is a property of systems with a bounded Hamiltonian
(spins, two-level optical media); a free gas, with unbounded kinetic
energy, has no negative-temperature states.
\end{mastery}

\begin{mastery}
\textbf{The Onsager reciprocal relations: symmetry near equilibrium}. The
entropy-generation accounting of this chapter charges each irreversible flow the
product of a flux and the thermodynamic force that drives it: a heat flux driven
by a temperature gradient, an electric current driven by a voltage, a diffusion
flux driven by a concentration gradient. Near equilibrium each flux $J_i$ is
linear in all the forces $X_j$, $J_i = \sum_j L_{ij} X_j$, and the entropy
generation is the sum of the products, $\dot S_{\text{gen}} = \sum_i J_i X_i \ge
0$. The off-diagonal coefficients $L_{ij}$ are the cross-effects: a temperature
gradient can drive an electric current (the thermoelectric effect the design
exercise uses), and a voltage can drive a heat flux (the Peltier effect the same
junction shows in reverse). Lars Onsager proved in 1931 that the matrix of these
coefficients is symmetric, $L_{ij} = L_{ji}$, so the coupling of force $j$ to
flux $i$ equals the coupling of force $i$ to flux $j$. The Seebeck coefficient
that turns a temperature difference into a voltage and the Peltier coefficient
that turns a current into a heat flux are therefore not independent numbers; they
are locked together by $L_{ij} = L_{ji}$, and a thermoelectric engineer who
measures one has measured the other. The proof rests on the time-reversal
symmetry of the microscopic equations of motion, the same symmetry the arrow-of-
time section wrestles with: the correlation of a fluctuation in quantity $i$ with
a later fluctuation in quantity $j$ equals the time-reversed correlation, and
that microscopic symmetry surfaces as the macroscopic reciprocity of the
transport coefficients. The reciprocal relations are the reason the coupled
irreversible processes of a real device, heat and charge in a thermoelectric,
heat and mass in a drying solid, obey a hidden economy: the cross-couplings come
in matched pairs, and the second law constrains the whole matrix to be positive,
so no combination of cross-effects can be arranged to lower the total entropy
generation below zero.
\end{mastery}

\section{Entropy in chemical and biological systems}\pathtag{standard}

Chemical reactions and biological processes both involve changes in
composition at approximately constant temperature and pressure. The
relevant state function is the Gibbs free energy $G = H - TS$, and
the change $\Delta G$ across a process at constant $T$ and $p$ has
two competing contributions: $\Delta H$ (the heat absorbed at
constant pressure) and $-T \Delta S$ (the negative of the
temperature times the entropy change).

The equilibrium condition for a chemical reaction at constant $T$
and $p$ is $\Delta G = 0$. The criterion for spontaneous direction is
$\Delta G < 0$. A reaction is spontaneous when either $\Delta H <
0$ (exothermic) and $\Delta S > 0$ (entropy-increasing) together, or
when one of the two opposes the other but is dominated.

\engterm{Worked example: water-gas-shift reaction}. The reaction
$\text{CO} + \text{H}_2\text{O} \rightleftharpoons \text{CO}_2 +
\text{H}_2$ at $25\,\degC$ has $\Delta H^\circ = -41.2\,\text{kJ}/
\text{mol}$ (exothermic) and $\Delta S^\circ = -42.1\,\text{J}/(
\text{mol}\cdot\si{\kelvin})$. At $T = 298\,\si{\kelvin}$,
$T \Delta S = -12.5\,\text{kJ}/\text{mol}$, giving $\Delta G^\circ =
-41.2 - (-12.5) = -28.7\,\text{kJ}/\text{mol}$: the forward direction
is thermodynamically favoured at room temperature. At higher
temperatures, the $T \Delta S$ term grows in magnitude (and is
negative), so $\Delta G$ rises and the reaction becomes less
favoured. At sufficient temperature, the reverse direction becomes
favoured. The temperature dependence of $\Delta G$ is the working
content of Le Chatelier's principle in temperature.
Figure~\ref{fig:vol04ch04:gibbs-temperature} plots $\Delta G^\circ(T)$
for this reaction as the straight line $\Delta H^\circ - T\Delta S^\circ$;
the crossover temperature $T^\ast = \Delta H^\circ/\Delta S^\circ \approx
978\,\si{\kelvin}$ is where the forward and reverse directions balance.
The script \texttt{code/reaction\_equilibrium.py} produces the line and
the equilibrium-constant sweep in \texttt{data/wgs\_equilibrium.csv}.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-gibbs-temperature.tex}

\engterm{Equilibrium constants}. For a reaction $aA + bB
\rightleftharpoons cC + dD$ at constant $T$ and $p$,
\[
\Delta G = \Delta G^\circ + RT \ln K,
\]
where $K = (a_C^c a_D^d)/(a_A^a a_B^b)$ is the reaction quotient in
activities. At equilibrium, $\Delta G = 0$ and $K = K_{\text{eq}} =
\exp(-\Delta G^\circ/RT)$. The temperature dependence of $K_{\text{eq}}$
is given by the \engterm{van't Hoff equation}
\[
\frac{d \ln K_{\text{eq}}}{dT} = \frac{\Delta H^\circ}{RT^2}.
\]
Exothermic reactions have $K_{\text{eq}}$ that decreases with
temperature; endothermic reactions have $K_{\text{eq}}$ that
increases. The engineering design of chemical reactors balances
temperature for kinetics (Arrhenius rate increases with $T$) against
equilibrium yield (which depends on the sign of $\Delta H$).
Volume~V develops the chemical-kinetic side.

\engterm{Biological systems}. A living cell maintains itself in a
state of low entropy relative to chemical equilibrium with its
environment. The maintenance requires a continuous flow of free
energy in (food, photons, chemical fuel) and a corresponding flow of
entropy out (heat, low-entropy waste products). The fundamental
energy currency in biology is the hydrolysis of ATP, which delivers
$\Delta G \approx -30\,\text{kJ}/\text{mol}$ at typical cellular
conditions. Biological energy accounting at the cell, organism, and
ecosystem level is the working topic of Volume~VI; the
thermodynamic foundation is the same Gibbs free-energy framework.

The second law does not forbid a cell from being more ordered than
its environment. The cell maintains its low-entropy state by
\emph{exporting} entropy at a rate at least as great as the rate at
which its internal organisation would otherwise dissipate. A
biological process whose entropy export to the environment is less
than the rate of internal disorder accumulation cannot be
maintained; the cell, the organism, or the ecosystem dies. The
second law applies to the cell-plus-environment, not to the cell
alone, and is satisfied through the entropy export.

The ATP currency repays a closer look, because the number a textbook quotes for
its hydrolysis is not the number the cell works with. The standard free energy of
ATP hydrolysis, $\text{ATP} + \text{H}_2\text{O} \to \text{ADP} +
\text{P}_\text{i}$, is $\Delta G^\circ \approx -30\,\text{kJ}/\text{mol}$ at
standard concentrations of one molar. A cell does not hold its metabolites at one
molar. The working free energy is the concentration-corrected value
\[
\Delta G = \Delta G^\circ + RT\ln\frac{[\text{ADP}][\text{P}_\text{i}]}
{[\text{ATP}]},
\]
and a cell keeps the ATP concentration high and the ADP and phosphate
concentrations low, so the logarithm is a large negative number that drives
$\Delta G$ well below its standard value, to roughly $-50$ to
$-60\,\text{kJ}/\text{mol}$ in a working cell. The cell holds its ATP-to-ADP
ratio far from chemical equilibrium precisely so that each hydrolysis delivers
more free energy than the standard figure promises. Maintaining that
disequilibrium is itself an entropy-exporting activity: the cell spends free
energy pumping the concentrations away from their equilibrium ratio, and the
entropy it exports in doing so is the price of keeping its energy currency
valuable. The distance of the ATP system from equilibrium, quantified by that
logarithm, is a stored exergy the cell draws on the way the concentration cell of
the worked examples draws on a concentration difference.

The activity correction that turns $\Delta G^\circ$ into the working $\Delta G$
carries past biology into every real chemical process. The equilibrium constant
was written in activities, not concentrations, and the distinction matters
whenever a solution is concentrated or an ion is charged. The activity is the
concentration times an activity coefficient $\gamma$, $a_i = \gamma_i c_i/c^\circ$,
and $\gamma$ departs from one because the species interact: ions feel their
neighbours' charges, and the Debye-Huckel theory gives the leading correction,
$\ln\gamma_\pm \propto -z_+ z_- \sqrt{I}$, with $I$ the ionic strength. An
engineer sizing a crystalliser, a battery electrolyte, or a scrubber who used
concentrations in place of activities would misplace the equilibrium, sometimes
by a large factor, because the entropy of a real solution is not the ideal
mixing entropy of section~4.1 but that entropy corrected for the ordering the
interactions impose. The activity coefficient is the thermodynamic bookkeeping of
that non-ideal ordering, and it is the reason handbook equilibrium data are
tabulated against ionic strength rather than as single numbers.

\section{Information entropy preview}\pathtag{standard}

The Shannon information entropy of a probability distribution $\{p_i\}$
is \cite{paper:shannon1948}
\[
H = -\sum_i p_i \log p_i,
\]
in units that depend on the logarithm base (bits for $\log_2$, nats
for $\log_e$). The formula coincides, up to the Boltzmann constant,
with the Gibbs entropy
\[
S = -k_B \sum_i p_i \ln p_i,
\]
where $\{p_i\}$ is the probability of microstate $i$ in a
statistical ensemble. The coincidence is not accidental. The Gibbs
entropy was Shannon's starting point, and the two quantities measure
the same thing: the average uncertainty (in microstates, or in
messages) corresponding to a known macrostate (or distribution).

The information-theoretic view of entropy provides several working
relations:

\begin{itemize}[leftmargin=*]
  \item \engterm{Landauer's principle}: erasing one bit of
    information at temperature $T$ dissipates at least $k_B T \ln 2$
    of energy as heat. The principle gives a thermodynamic lower
    bound on the energy cost of irreversible computation; it is
    experimentally verified to within the limits of contemporary
    measurement.
  \item \engterm{Maxwell's demon resolved}: a demon that sorts fast
    and slow molecules to produce a temperature difference must
    measure each molecule's velocity and store the result, then
    eventually erase the stored information. The erasure dissipates
    energy at the Landauer bound; the bound exactly compensates the
    work that the demon could extract from the temperature
    difference. Maxwell's demon does not violate the second law; the
    demon plus its memory satisfies the second law together
    \cite{paper:shannon1948}.
  \item \engterm{Mutual information}: the entropy reduction in one
    system due to knowledge of another. Engineering applications
    range from communication theory (the channel capacity is the
    maximum mutual information) to feedback control (the optimal
    controller's performance is bounded by the information rate of
    the sensor).
\end{itemize}

\engterm{Relative entropy and the maximum-entropy principle}. The Shannon
entropy measures the uncertainty in one distribution; the \engterm{relative
entropy} measures the distance from one distribution to another. For a
distribution $\{p_i\}$ compared to a reference $\{m_i\}$, the relative entropy,
or Kullback-Leibler divergence, is
\[
D(p\,\|\,m) = \sum_i p_i \ln\frac{p_i}{m_i}.
\]
It is never negative, and it is zero only when $p$ and $m$ coincide, which follows
from the convexity of $x\ln x$ by Jensen's inequality: the average of the
convex function lies above the function of the average, and the gap is exactly
$D$. Relative entropy is not symmetric, $D(p\,\|\,m) \ne D(m\,\|\,p)$ in general,
so it is a directed measure of the information gained on learning that the true
distribution is $p$ when one had assumed $m$, quoted in nats for the natural
logarithm and in bits for base two. When the reference is uniform over $W$ states,
$m_i = 1/W$, the divergence reduces to $D = \ln W - H$, the shortfall of the
distribution's entropy below the maximum a uniform distribution would carry, so
relative entropy and Shannon entropy are two readings of the same quantity against
different baselines. Figure~\ref{fig:vol04ch04:relative-entropy} draws both for a
biased coin: the entropy peaks where the divergence from uniform vanishes, and the
two curves are mirror images summing to one bit.

The relative entropy grounds the \engterm{maximum-entropy principle}, the working
rule for assigning a probability distribution from incomplete information. Faced
with a set of constraints, typically the normalisation and a few measured average
values, the principle selects the distribution of largest Shannon entropy
consistent with them, equivalently the distribution closest to uniform in the
relative-entropy sense. Any other choice would smuggle in information the
constraints do not supply. Carrying out the maximisation with Lagrange multipliers,
exactly the calculation the derivation exercise performs for fixed average energy,
returns the exponential family: with a single energy constraint the maximum-entropy
distribution is the Boltzmann distribution $p_i \propto e^{-\beta E_i}$, and the
multiplier $\beta$ is the inverse temperature. The canonical ensemble is therefore
not a physical postulate but the least-committed distribution consistent with a
known average energy, which is why statistical mechanics and inference share one
piece of machinery. An engineer estimating a spectrum from a few moments, a
signal from a few samples, or a reliability distribution from a few failure
statistics reaches for the same maximum-entropy construction, and the entropy it
maximises is the Shannon entropy that coincides with the thermodynamic one.

The Shannon entropy measures the uncertainty of a random source averaged over its
distribution, and a second notion of information content asks a sharper question
about a single object rather than a distribution. The \engterm{algorithmic
complexity}, or Kolmogorov complexity, of a string is the length of the shortest
program that produces it, a measure of how compressible that one string is with no
probability distribution assumed. A random-looking bit string of length $n$ has
algorithmic complexity near $n$, because nothing shorter than the string itself
generates it, while the first million digits of $\pi$, which look random by every
statistical test, have tiny algorithmic complexity, because a short program prints
them all. The two measures agree on average: for a string drawn from a known
source, the expected algorithmic complexity per symbol converges to the Shannon
entropy of the source, so a data compressor cannot on average pack a stream below
its Shannon entropy, and a good compressor approaches that floor. The measures
part company on the individual object, where Shannon entropy is silent and
algorithmic complexity is defined but uncomputable, since no procedure can find the
shortest program for an arbitrary string. The distinction matters to an engineer
in two places: it is the reason lossless compression has a hard floor set by the
source entropy no algorithm can beat, and it is the reason a stream that passes
every randomness test may still be the deterministic output of a short seed, the
gap a cryptographer exploits and a careless designer of a random-number generator
falls into. Both measures rest on the same $-\log p$ surprisal that the
thermodynamic entropy counts, and both descend to the Boltzmann count when the
object is a microstate and the distribution is the ensemble.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-relative-entropy.tex}

\begin{mastery}
\textbf{The Szilard engine: one bit is worth $k_B T \ln 2$ of work}. Leo Szilard's
1929 thought engine strips Maxwell's demon down to a single molecule and makes the
exchange rate between information and work exact. A box of volume $V$ holds one
molecule in contact with a bath at temperature $T$. A partition drops in the
middle, and a measurement records which half the molecule is in, one bit of
information. Knowing the side, the operator inserts a piston on the empty half and
lets the single-molecule gas expand isothermally against it back to the full
volume. The work the expanding one-molecule gas delivers is the isothermal-expansion
integral $W = \int_{V/2}^{V} p\,dV = k_B T \int_{V/2}^{V} dV/V = k_B T \ln 2$,
drawn as heat from the bath and converted entirely to work. One bit of information
about the molecule's location has been converted into $k_B T \ln 2$ of work, the
same $k_B T \ln 2$ that appears in the Landauer bound, and the coincidence is not
accidental: they are the two directions of one exchange rate between information and
energy. The engine seems to draw work from a single reservoir in a cycle, the
Kelvin-Planck violation, until the memory is accounted for. The measurement wrote a
bit into the operator's memory, and to return the whole apparatus, memory included,
to its starting state that bit must be erased, which by Landauer costs at least
$k_B T \ln 2$ of heat dumped to the bath. The erasure cost exactly cancels the work
extracted, and the cycle delivers zero net work over the full round trip, memory
reset included. The Szilard engine is the cleanest statement in the chapter that
information is physical: a bit has a thermodynamic price, the demon and its memory
obey the second law together, and the ledger balances only when the cost of
forgetting is entered alongside the value of knowing.
\end{mastery}

Volume~VII develops information theory in its own right; Volume~XII
returns to the information-entropy view in the context of computation
and complexity. For the present chapter, the engineering message is
that thermodynamic entropy and information entropy are the same
quantity in different working units.

\begin{mastery}
\textbf{Gibbs entropy, Boltzmann entropy, and the cost of coarse-graining}. Two
entropies bear these names, and keeping them apart resolves a paradox that
otherwise makes the second law look inconsistent with mechanics. The
\engterm{fine-grained Gibbs entropy} is $S_G = -k_B\int \rho\ln\rho\,d\Gamma$
computed over the full phase-space density $\rho$ of an isolated system. Liouville's
theorem, the statement that the phase-space density flows like an incompressible
fluid under the equations of motion, makes this quantity exactly constant in
time: an isolated system's fine-grained entropy never changes at all. That is not
the second law, and if it were the only entropy there would be no arrow of time.
The \engterm{coarse-grained Boltzmann entropy} is the one the second law governs.
Partition phase space into cells of finite resolution, the resolution set by what
the observer can actually measure, and assign each macrostate the entropy $S_B =
k_B \ln W$ of the phase-space volume its cells occupy. As the true density $\rho$
develops the fine, whorled filaments that Liouville's theorem demands, keeping
its fine-grained entropy fixed, those filaments spread across ever more cells of
the coarse partition, and the coarse-grained volume swells. The Boltzmann
entropy rises even though the Gibbs entropy is frozen, and the difference is the
information the coarse-graining discards: the filaments carry the memory of the
initial condition down to arbitrarily fine scale, and the observer who cannot
resolve them measures an entropy increase. The relative entropy of section~4.4 is
the exact bookkeeper of the gap, $S_B - S_G = k_B D(\rho\,\|\,\rho_{\text{coarse}})
\ge 0$, the divergence between the true density and its coarse-grained image. The
lesson closes the arrow-of-time discussion from the measurement side: entropy
increase is not a property of the microscopic dynamics, which are reversible and
conserve the fine-grained entropy, but of the coarse description an observer is
forced to use, and the rate of increase is set by how fast the dynamics stir
fine structure past the observer's resolution. The engineer never sees this
subtlety because the macroscopic variables fix the coarse-graining once and for
all; the paradox lives only for someone who imagines tracking every molecule.
\end{mastery}

\section{Free energy and equilibrium}\pathtag{core}

A system at fixed temperature and pressure evolves toward the state
that minimises its Gibbs free energy $G = H - TS$. The equilibrium
state is the one for which $dG = 0$ at constant $T$ and $p$. For a
system that can exchange matter with its surroundings or undergo
chemical reaction, the variational condition is
\[
dG = -S \,dT + V \,dp + \sum_i \mu_i \,dN_i = 0,
\]
with the $\mu_i$ chemical potentials acting as the conjugate forces
to the composition variables $N_i$.

The chemical-potential balance gives the working equilibrium
conditions:

\begin{itemize}[leftmargin=*]
  \item \engterm{Phase equilibrium}: the chemical potential of any
    species must be the same in every phase that contains it. The
    Clausius-Clapeyron relation of chapter~2 is the temperature-
    pressure form of this condition for a pure substance.
  \item \engterm{Chemical equilibrium}: for a reaction $\sum_i \nu_i
    A_i = 0$, the equilibrium condition is $\sum_i \nu_i \mu_i = 0$,
    which converts via the standard-state expansion to $K_{\text{eq}}
    = \exp(-\Delta G^\circ/RT)$.
  \item \engterm{Diffusional equilibrium}: at equilibrium, the
    chemical potential is uniform across the system, even when the
    concentration is not (a system in a gravitational or electric
    field may have a concentration gradient in equilibrium, with
    the gradient set by the gradient of the external potential).
\end{itemize}

The phase-equilibrium condition earns a derivation, because equating chemical
potentials across a coexistence line yields the Clausius-Clapeyron relation that
governs every boiling, melting, and sublimation an engineer must place on a phase
diagram. Along the line where liquid and vapour coexist, the two chemical
potentials are equal at every point, $\mu_{\text{liq}}(T,p) =
\mu_{\text{vap}}(T,p)$. Moving along the line by $dT$ and $dp$ keeps them equal,
so their differentials match, $d\mu_{\text{liq}} = d\mu_{\text{vap}}$. Each
differential is $d\mu = -s\,dT + v\,dp$ in molar entropy $s$ and molar volume
$v$, the Gibbs-Duhem form of the potential's dependence, so equality gives
$-s_{\text{liq}}\,dT + v_{\text{liq}}\,dp = -s_{\text{vap}}\,dT +
v_{\text{vap}}\,dp$, which rearranges to
\[
\frac{dp}{dT} = \frac{s_{\text{vap}} - s_{\text{liq}}}{v_{\text{vap}} -
v_{\text{liq}}} = \frac{\Delta s_{\text{vap}}}{\Delta v_{\text{vap}}}
= \frac{L}{T\,\Delta v},
\]
using $\Delta s = L/T$ for the reversible phase change at temperature $T$. The
slope of the coexistence line is the entropy of the transition over its volume
change, a purely thermodynamic statement with no adjustable constant. For
vaporization the vapour volume dwarfs the liquid's and the vapour is nearly
ideal, $\Delta v \approx v_{\text{vap}} = RT/p$, which turns the relation into
$d\ln p/dT = L/(RT^2)$, the integrated form plotted in
figure~\ref{fig:vol04ch04:clausius-clapeyron}. The chemical-potential equality
is the root, the Clausius-Clapeyron slope is the fruit, and the straight line of
$\ln p$ against $1/T$ is the second law's signature written on a vapour-pressure
chart: its slope is the entropy of vaporization the phase equilibrium holds in
balance.

\engterm{The exergy concept}. The Gibbs free energy of a system
relative to a specified reference state (typically the ambient at
$T_0$, $p_0$, and the ambient atmospheric composition) is called the
\engterm{exergy} or \engterm{availability}. The exergy is the maximum
useful work that can be extracted from the system in reaching
equilibrium with the reference state. Exergy is path-independent (a
state function) but reference-dependent (the choice of $T_0$, $p_0$
matters).

Exergy analysis is the second-law-based working tool for evaluating
energy systems. Each component of a process flow-sheet has an
\engterm{exergy destruction rate} equal to $T_0 \dot{S}_{\text{gen}}$,
where $\dot{S}_{\text{gen}}$ is the component's entropy generation
rate from chapter~1. The total exergy destruction equals the gap
between the actual work output and the maximum work output (Carnot
or its analogues). Exergy analysis identifies which components in a
process destroy the most exergy and, therefore, where redesign would
deliver the largest second-law efficiency gain.
Figure~\ref{fig:vol04ch04:exergy-flow} is the Grassmann diagram of a
simple steam plant, with the component exergy destructions drawn as the
side-arrows; the boiler dominates, which an energy balance alone would
not reveal. The temperature-entropy diagram of
figure~\ref{fig:vol04ch04:ts-diagram-carnot} gives the reversible
reference against which exergy destruction is measured: the area
$(T_H - T_C)\Delta S$ enclosed by a Carnot rectangle is the maximum
work, and any real cycle falls short of it by the total
$T_0 \dot S_{\text{gen}}$ summed over components. The accounting in
\texttt{code/exergy\_balance.py} reproduces the diagram's percentages.

\begin{mastery}
\textbf{The Gouy-Stodola theorem}. The link between entropy generation
and lost work is not a heuristic; it is an identity. Consider any
steady-flow device that exchanges heat only with a single environment at
temperature $T_0$. An energy balance and an entropy balance written
across the device combine into a single statement: the work the device
actually delivers falls short of the reversible work by exactly
$T_0$ times the rate at which the device generates entropy,
\[
\dot W_{\text{lost}} = \dot W_{\text{rev}} - \dot W_{\text{actual}}
= T_0\,\dot S_{\text{gen}}.
\]
The result is the Gouy-Stodola theorem, named for Georges Gouy and
Aurel Stodola, who stated it independently around the turn of the
twentieth century. The exergy destroyed in a component is $T_0
\dot S_{\text{gen}}$ because that is the work the component could have
delivered and did not. The theorem turns the abstract injunction
``minimise entropy generation'' into a money quantity: every
$\si{\watt\per\kelvin}$ of entropy a component generates is $T_0$ watts
of shaft power, fuel, or refrigeration the plant will never recover.
For a plant rejecting heat to a $T_0 = 290\,\si{\kelvin}$ environment,
one $\si{\watt\per\kelvin}$ of avoidable entropy generation is $290$
watts of lost capacity, charged at the price of whatever the plant
sells. The theorem also fixes the target of design: a component cannot
be improved past $\dot S_{\text{gen}} = 0$, and the gap between its
current generation and zero is the exact ceiling on what any redesign of
that component can recover. Entropy-generation minimisation, exergy
analysis, and second-law efficiency are three readings of this one
identity.
\end{mastery}

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-ts-diagram-carnot.tex}

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-exergy-flow.tex}

\section{The thermodynamic potentials and the Maxwell relations}\pathtag{core}

The free-energy criteria of the previous section name $F$ and $G$ as the
quantities a system minimises at fixed $(T,V)$ and fixed $(T,p)$, but they
treat the potentials as given. The potentials follow from a single
differential, the combined statement of the first and second laws, by a
construction worth carrying in full, because the same construction generates
every standard thermodynamic identity and the Maxwell relations that turn a
hard-to-measure derivative into an easy one \cite[ch.~5]{text:v6c02-callen-thermodynamics-1985}.

\engterm{The Gibbs equation}. The first law for a closed simple compressible
substance writes the change in internal energy as heat in minus work out,
$dU = \delta Q - \delta W$. For a reversible path the heat is $\delta Q_{\text{rev}}
= T\,dS$ by the definition of entropy, and the work is $\delta W_{\text{rev}} =
p\,dV$. Substituting both gives
\[
dU = T\,dS - p\,dV,
\]
the \engterm{Gibbs equation}, also called the first $T\,dS$ equation. It was
derived along a reversible path, but it relates state functions only, $U$, $S$,
and $V$, so it holds between any two equilibrium states regardless of the path
that connects them. The reversible path is the scaffold; the result is
path-independent. The Gibbs equation is the master differential of equilibrium
thermodynamics, and the three potentials beyond $U$ are its Legendre transforms.

\engterm{Legendre transforms and the natural variables}. The internal energy in
the form $U(S,V)$ has the entropy and the volume as its \engterm{natural
variables}: from $dU = T\,dS - p\,dV$ we read the partial derivatives directly,
\[
T = \left(\pdv{U}{S}\right)_V, \qquad
p = -\left(\pdv{U}{V}\right)_S.
\]
The natural variables of $U$ are awkward for the laboratory and the plant, where
the controlled quantities are temperature and pressure, not entropy and volume.
A \engterm{Legendre transform} swaps a natural variable for the partial
derivative conjugate to it, subtracting the product of the conjugate pair so that
the new function's differential carries the new variable. To trade the volume $V$
for the pressure $p = -(\partial U/\partial V)_S$, add $pV$:
\[
H \equiv U + pV, \qquad
dH = dU + p\,dV + V\,dp = T\,dS + V\,dp,
\]
the \engterm{enthalpy} $H(S,p)$, whose natural variables are entropy and
pressure. To trade the entropy $S$ for the temperature $T = (\partial
U/\partial S)_V$, subtract $TS$:
\[
F \equiv U - TS, \qquad
dF = dU - T\,dS - S\,dT = -S\,dT - p\,dV,
\]
the \engterm{Helmholtz free energy} $F(T,V)$. Doing both transforms gives the
\engterm{Gibbs free energy}
\[
G \equiv U - TS + pV = H - TS, \qquad
dG = -S\,dT + V\,dp,
\]
with temperature and pressure as its natural variables, the two quantities a
bench experiment or a steady-flow plant most often holds fixed. The four
potentials and the transforms linking them are collected in
figure~\ref{fig:vol04ch04:thermodynamic-potentials}. Each potential is the
energy minimised at equilibrium when its own natural variables are held constant,
which is why the choice of potential follows the choice of constraint: $U$ at
fixed $S$ and $V$, $H$ at fixed $S$ and $p$, $F$ at fixed $T$ and $V$, $G$ at
fixed $T$ and $p$.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-thermodynamic-potentials.tex}

\engterm{The Maxwell relations}. Each potential's differential has the form
$d\Phi = X\,da + Y\,db$, from which $X = (\partial \Phi/\partial a)_b$ and $Y =
(\partial \Phi/\partial b)_a$. A potential is a state function, so its mixed
second partial derivatives are equal regardless of the order of differentiation,
$\partial^2\Phi/\partial a\,\partial b = \partial^2\Phi/\partial b\,\partial a$,
which forces
\[
\left(\pdv{X}{b}\right)_a = \left(\pdv{Y}{a}\right)_b.
\]
Applied to the four differentials in turn, this equality of mixed partials yields
the four \engterm{Maxwell relations}:
\begin{align*}
\text{from } dU:\quad & \left(\pdv{T}{V}\right)_S = -\left(\pdv{p}{S}\right)_V, &
\text{from } dH:\quad & \left(\pdv{T}{p}\right)_S = \left(\pdv{V}{S}\right)_p, \\
\text{from } dF:\quad & \left(\pdv{S}{V}\right)_T = \left(\pdv{p}{T}\right)_V, &
\text{from } dG:\quad & \left(\pdv{S}{p}\right)_T = -\left(\pdv{V}{T}\right)_p.
\end{align*}
The relations earn their place because they trade a derivative of entropy, which
no instrument reads directly, for a derivative of pressure, volume, and
temperature, which an instrument reads at once. The third relation says the
entropy's response to volume at fixed temperature equals the pressure's response
to temperature at fixed volume, $(\partial S/\partial V)_T = (\partial
p/\partial T)_V$, and the right side is measured by warming a sealed rigid vessel
and watching the pressure climb. The fourth relation converts the entropy's
pressure dependence into the thermal expansion coefficient. Every table of
entropy, enthalpy, and free energy in the engineering handbooks is built by
integrating measured $p$-$V$-$T$ data through these relations; the Maxwell
relations are the bridge from what a gauge reads to what a thermodynamic ledger
needs.

\engterm{Exergy from the combined first and second law}. The exergy named in the
previous section follows from the same Gibbs equation applied to a system plus its
environment. Consider a closed system at state $(T,p)$ that will come to
equilibrium with an environment fixed at $(T_0,p_0)$. The maximum work the system
can deliver is its work output less the work it must spend pushing back the
atmosphere, and the second law caps it. Combining an energy balance, $dU = \delta
Q - \delta W$, with the entropy bound on the heat exchanged with the environment,
$\delta Q \le T_0\,dS$ when the only heat reservoir is the environment, gives the
useful-work bound
\[
\delta W_{\text{useful}} \le -\,d\!\left(U + p_0 V - T_0 S\right).
\]
The quantity in parentheses is the \engterm{availability function}, and the
exergy of the system relative to the dead state is its drop from the current
state to equilibrium with the environment,
\[
\Phi = (U - U_0) + p_0(V - V_0) - T_0(S - S_0),
\]
the maximum useful work extractable in reaching the dead state. The combined-law
derivation makes plain why exergy is reference-dependent: the $T_0$ and $p_0$ that
weight the entropy and volume terms are the environment's, not the system's, and
a different environment yields a different ceiling. The exergy destruction
$T_0\,\dot S_{\text{gen}}$ of the Gouy-Stodola theorem is the rate at which a real
process erodes this availability, the gap between the bound and the actual work.

\engterm{The third law and the behaviour of heat capacity as $T \to 0$}. The
Gibbs equation fixes entropy differences; it does not fix the absolute entropy,
because integrating $dS = \delta Q_{\text{rev}}/T$ leaves a constant of
integration. The \engterm{third law of thermodynamics}, in Nernst's and Planck's
form, fixes that constant: the entropy of every system in internal equilibrium
approaches the same finite value, taken as zero, as the temperature approaches
absolute zero. The statistical reading is immediate from $S = k_B \ln W$: a system
with a unique, non-degenerate ground state has $W \to 1$ and $S \to 0$ as $T \to
0$, because all the thermal excitations empty out and only the single ground
configuration remains, exactly as the Einstein-solid count of the worked-examples
section delivered.

The third law constrains the low-temperature heat capacity. The absolute entropy
is the integral of the heat capacity weighted by $1/T$,
\[
S(T) = \int_0^T \frac{c_p(T')}{T'}\,dT',
\]
and this integral converges to a finite $S(0)$ only if $c_p(T') \to 0$ at least
as fast as $T'$ as $T' \to 0$, since otherwise the $1/T'$ weight makes the
integrand non-integrable at the lower limit. A material with a non-vanishing heat
capacity at absolute zero would carry infinite entropy, which the count forbids.
The heat capacity must therefore vanish at absolute zero, and the way it vanishes
is set by the spectrum of low-energy excitations: a crystalline solid follows the
Debye $c_v \propto T^3$ law as its phonon modes freeze out, while the conduction
electrons of a metal add a term linear in $T$.
Figure~\ref{fig:vol04ch04:third-law-cp} plots the heat capacity climbing from
zero at absolute zero toward the classical Dulong-Petit plateau, the curve whose
low-temperature tail the third law demands. A second consequence is the
unattainability of absolute zero: because every entropy-lowering step yields a
smaller temperature drop as $T \to 0$, no finite sequence of reversible
operations reaches absolute zero, the operational form of the third law that
bounds every cryogenic process.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-third-law-cp.tex}

\begin{mastery}
\textbf{The Maxwell relations as a measurement lever}. The working power of the
Maxwell relations is that they let an engineer compute a quantity no instrument
reads from quantities every instrument reads. Take the question of how the
internal energy of a real gas depends on its volume at fixed temperature,
$(\partial U/\partial V)_T$, the quantity that decides whether a gas heats or
cools on free expansion. Start from the Gibbs equation written per unit and
divide by $dV$ at fixed $T$,
\[
\left(\pdv{U}{V}\right)_T = T\left(\pdv{S}{V}\right)_T - p,
\]
then substitute the third Maxwell relation $(\partial S/\partial V)_T = (\partial
p/\partial T)_V$ to obtain the \engterm{thermodynamic equation of state}
\[
\left(\pdv{U}{V}\right)_T = T\left(\pdv{p}{T}\right)_V - p.
\]
The right side is built entirely from the measurable $p$-$V$-$T$ surface. For an
ideal gas $p = Nk_B T/V$ gives $T(\partial p/\partial T)_V = p$, so $(\partial
U/\partial V)_T = 0$: the internal energy of an ideal gas is independent of
volume, which is why its free expansion leaves the temperature unchanged, the
result asserted without proof in section~4.1. For a van der Waals gas the same
substitution gives $(\partial U/\partial V)_T = a/V_m^2 > 0$, the attractive-term
energy that makes a real gas cool on free expansion and underwrites the
Joule-Thomson refrigeration of the gas-liquefaction industry. One Maxwell
relation converts an unmeasurable energy derivative into the slope of a pressure
gauge against a thermometer.
\end{mastery}

\section{An end-to-end case study: the second-law audit of a steam plant}\pathtag{standard}

The exergy machinery earns its place when an engineer must find where a
working plant wastes its fuel and decide where a fixed redesign budget
will pay back most. A first-law energy balance answers the first
question wrongly often enough that the second-law audit is worth the
extra arithmetic. We carry a single subcritical steam plant through a
complete exergy accounting, component by component, and show that the
component the energy balance blames is not the component the exergy
balance indicts.

\subsection{The plant and its operating point}

The plant is a coal-fired subcritical Rankine unit of modest size. We
fix its operating point with round numbers that a textbook cycle would
recognise and that make the arithmetic legible. The boiler burns fuel
whose chemical exergy enters at the rate $\dot X_{\text{fuel}} =
100\,\text{MW}$; we normalise every later quantity to this figure, so
the audit reads directly as percentages of fuel exergy in. The steam
leaves the boiler at $T_3 = 810\,\si{\kelvin}$ ($\approx 537\,\degC$)
and enters the turbine; the condenser holds the cold end at $T_1 =
311\,\si{\kelvin}$ ($\approx 38\,\degC$), set by the cooling water. The
environment, the dead state against which exergy is reckoned, sits at
$T_0 = 290\,\si{\kelvin}$ and $p_0 = 1\,\text{atm}$. The plant delivers
a measured net electrical output of $\dot W_{\text{net}} = 40\,\text{MW}$,
so its first-law thermal efficiency is $\eta_I = 40/100 = 0.40$, an
unremarkable and realistic number for such a unit.

The flame in the furnace reaches an adiabatic combustion temperature of
order $T_{\text{flame}} \approx 1900\,\si{\kelvin}$, while the working
fluid it heats never exceeds $810\,\si{\kelvin}$. That gap, between what
the fuel could in principle deliver and the temperature at which the
steam actually receives it, is where most of the plant's exergy will
turn out to be destroyed, and an energy balance cannot see it at all,
because energy crosses the gap conserved.

\subsection{Where the first law looks}

A first-law balance tracks energy. Of the $100\,\text{MW}$ of fuel
energy in, $40\,\text{MW}$ leaves as net work and the remaining
$60\,\text{MW}$ must leave as heat, almost all of it through the
condenser to the cooling water. The energy ledger therefore reads: a
small stack loss of perhaps $6$ to $8\,\text{MW}$ in the flue gas, a
negligible mechanical loss, and a dominant $\approx 52\,\text{MW}$
dumped at the condenser. Read literally, the energy balance says the
condenser is the plant's great waste, and an engineer who trusted it
would spend the redesign budget reclaiming condenser heat. The blue bars
of figure~\ref{fig:vol04ch04:exergy-destruction-bars} are this ledger:
the condenser towers over every other component.

The condenser heat is real, but it is heat rejected at $311\,\si{\kelvin}$,
barely above the $290\,\si{\kelvin}$ dead state. Its capacity to do work
is almost nil. Reclaiming it yields low-grade warmth fit for district
heating at best, not shaft power. The energy ledger counts joules; it
does not weight them by their ability to become work, and that omission
is what sends the redesign budget to the wrong component.

\subsection{Where the second law looks}

The exergy audit weights every flow by its distance from the dead state
and charges each component the exergy it destroys, $T_0\,\dot S_{\text{gen}}$.
We take the components in turn.

\paragraph{Boiler and combustion.} The heat $\dot Q_{\text{in}}$ added
to the steam is delivered from combustion gases near $T_{\text{flame}}$
into working fluid that averages, across the economiser, evaporator, and
superheater, a thermodynamic mean temperature of roughly
$\bar T_{\text{steam}} \approx 560\,\si{\kelvin}$. The heat added is the
fuel energy less the stack loss, about $\dot Q_{\text{in}} \approx
93\,\text{MW}$. The entropy generated by passing this heat down from the
flame to the steam is
\[
\dot S_{\text{gen,boiler}} \approx \dot Q_{\text{in}}
\left(\frac{1}{\bar T_{\text{steam}}} - \frac{1}{T_{\text{flame}}}\right)
= 93\times 10^6\left(\frac{1}{560} - \frac{1}{1900}\right)
= 93\times 10^6 \times (1.79 - 0.53)\times 10^{-3},
\]
which is $\dot S_{\text{gen,boiler}} \approx 1.17\times 10^5\,
\si{\watt\per\kelvin}$. By Gouy-Stodola the exergy destroyed is
$T_0\,\dot S_{\text{gen}} = 290 \times 1.17\times10^5 \approx
34\,\text{MW}$, with a further several megawatts destroyed by the
chemical irreversibility of combustion itself and by friction in the
flue path, bringing the boiler island to about $44\,\text{MW}$, or
$44\%$ of fuel exergy. The boiler, invisible to the energy balance, is
the plant's largest exergy sink.

\paragraph{Turbine.} A real turbine expands the steam with an isentropic
efficiency near $0.88$, so it generates entropy through the difference
between the actual and the ideal expansion. The exergy this destroys is
the gap between the work an isentropic turbine would deliver and the work
the real machine delivers, about $5\,\text{MW}$, or $5\%$ of fuel
exergy. The turbine is a well-developed component; its second-law
inefficiency is modest.

\paragraph{Condenser.} Here the two ledgers part company most sharply.
The condenser rejects $\approx 52\,\text{MW}$ of heat, the energy
balance's headline figure. But it rejects that heat at $311\,\si{\kelvin}$
into cooling water near the $290\,\si{\kelvin}$ dead state, across a gap
of only about $20\,\si{\kelvin}$. The entropy it generates is
\[
\dot S_{\text{gen,cond}} \approx \dot Q_{\text{rej}}
\left(\frac{1}{T_{\text{cw}}} - \frac{1}{T_{\text{cond}}}\right)
\approx 52\times10^6\left(\frac{1}{295} - \frac{1}{311}\right)
\approx 9.1\times10^3\,\si{\watt\per\kelvin},
\]
so the exergy destroyed is $290 \times 9.1\times10^3 \approx
2.6\,\text{MW}$, with the exergy that physically leaves with the warmed
cooling water adding a few more megawatts. The condenser island lands
near $3\%$ of fuel exergy destroyed plus a small exergy outflow, a tenth
of what it rejects as energy.

\paragraph{Feed pump.} The pump raises the condensate pressure with a
small work input and a small irreversibility, well under $1\%$ of fuel
exergy. It is included for completeness; it never governs.

\paragraph{Stack.} The flue gas leaves above ambient and carries both
energy and exergy out the stack, roughly $7\%$ of fuel exergy as a
combined thermal-exergy loss. This is a genuine second-law loss the
energy balance also records, and it is the one place the two ledgers
agree.

\subsection{The two ledgers side by side}

Figure~\ref{fig:vol04ch04:exergy-destruction-bars} sets the energy ledger
and the exergy ledger against each other component by component. The
condenser leads the energy ledger and nearly vanishes from the exergy
ledger; the boiler leads the exergy ledger and nearly vanishes from the
energy ledger. The Grassmann diagram of
figure~\ref{fig:vol04ch04:exergy-flow} is the same audit drawn as a flow:
fuel exergy enters at $100\%$, the boiler bleeds off its large share, the
turbine and condenser take their smaller cuts, and $40\%$ survives as net
work. The closing balance reads:
\[
\underbrace{40\%}_{\text{net work}} +
\underbrace{44\%}_{\text{boiler}} +
\underbrace{5\%}_{\text{turbine}} +
\underbrace{3\%}_{\text{condenser}} +
\underbrace{7\%}_{\text{stack}} +
\underbrace{1\%}_{\text{pump, misc}}
= 100\%,
\]
the exergy in accounted for, term by term, as either delivered work or
destruction.

\subsection{Second-law efficiency and the verdict}

The first-law efficiency, $\eta_I = 0.40$, compares work out to energy
in and treats every joule of fuel as equally valuable. The second-law
efficiency compares work out to \emph{exergy} in, the work the fuel could
in principle deliver. For a fuel-fired plant the fuel's chemical exergy
nearly equals its heating value, so here $\eta_{II} = \dot W_{\text{net}}/
\dot X_{\text{fuel}} = 40/100 = 0.40$ as well; the two efficiencies
coincide because we normalised to fuel exergy. The audit's payoff is not
the headline efficiency but the breakdown of the $60\%$ that does not
become work: $44$ points of it are destroyed in the boiler, only $3$ in
the condenser. The largest single recoverable loss is the temperature
gap between the flame and the steam. The design responses follow from the
ranking: raise the steam temperature and pressure toward supercritical
conditions to shrink the flame-to-steam gap, add feedwater heaters that
preheat the condensate with bled steam so that less heat must cross the
widest part of the gap, and add reheat to keep the expansion at high
average temperature. Each of these attacks the boiler's share, where the
exergy actually goes. Reinsulating the condenser, the move the energy
balance recommends, would recover warmth worth almost nothing in work.

The audit's lesson generalises past this one plant. An energy balance
tells the engineer how much energy leaves and where; it cannot tell the
engineer how much of that energy mattered, because it weights a joule at
$1900\,\si{\kelvin}$ the same as a joule at $311\,\si{\kelvin}$. The
exergy balance weights each joule by its distance from the dead state and
so points at the components where the work potential is destroyed. The
discipline of the second-law audit is to follow the exergy, not the
energy, when deciding where to spend the redesign budget. The
\texttt{code/exergy\_balance.py} script carries the full component
arithmetic and reproduces both ledgers of
figure~\ref{fig:vol04ch04:exergy-destruction-bars}.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-exergy-destruction-bars.tex}

\section{A second end-to-end case study: the second-law audit of a refrigerator}\pathtag{standard}

The steam plant turned heat into work and lost most of its fuel in the
boiler. A refrigerator runs the same accounting in reverse: it spends
work to pump heat uphill, and the question the second law answers is
where that purchased work goes. The first-law view of a refrigerator is
almost content-free, because in steady state every watt of electrical
work the compressor draws leaves as heat at the condenser along with the
heat lifted from the cold space; the energy ledger balances and says
nothing about quality. The exergy ledger does the work the first law
cannot, and it indicts components the energy balance never names. We
carry a single household-scale vapour-compression cycle through a
complete component audit, refrigerant state by refrigerant state, and
find that the cheap brass throttling valve destroys nearly as much work
potential as the compressor.

\subsection{The cycle and its operating point}

The unit is a domestic refrigerator holding a cold space at $T_C =
268\,\si{\kelvin}$ ($\approx -5\,\degC$) in a kitchen at $T_H =
303\,\si{\kelvin}$ ($\approx 30\,\degC$). The dead state, against which
exergy is reckoned, is the kitchen air at $T_0 = 303\,\si{\kelvin}$ and
$p_0 = 1\,\text{atm}$; the cold space lies below the dead state, which is
the feature that makes the audit read differently from the steam plant's.
The cabinet leaks heat in at a steady $\dot Q_C = 120\,\text{W}$, the
cooling load the cycle must lift. The compressor draws an electrical
input of $\dot W_{\text{in}} = 90\,\text{W}$ at the operating point, so
the measured coefficient of performance is $\mathrm{COP} = \dot Q_C/\dot
W_{\text{in}} = 120/90 = 1.33$. The refrigerant cycles through four
components in the order compressor, condenser, expansion valve,
evaporator, the states numbered $1$ through $4$ of
figure~\ref{fig:vol04ch04:refrigeration-ts}.

\subsection{The reversible benchmark}

A reversible refrigerator working between the same two temperatures has
the Carnot coefficient of performance
\[
\mathrm{COP}_{\text{Carnot}} = \frac{T_C}{T_H - T_C}
= \frac{268}{303 - 268} = \frac{268}{35} = 7.7.
\]
The reversible work to lift the $120\,\text{W}$ load is therefore $\dot
W_{\text{rev}} = \dot Q_C/\mathrm{COP}_{\text{Carnot}} = 120/7.7 =
15.6\,\text{W}$. The real machine spends $90\,\text{W}$ to do what
$15.6\,\text{W}$ could in principle do, so its second-law efficiency is
\[
\eta_{II} = \frac{\dot W_{\text{rev}}}{\dot W_{\text{in}}}
= \frac{15.6}{90} = 0.17,
\]
and the difference, $\dot W_{\text{in}} - \dot W_{\text{rev}} = 90 - 15.6
= 74\,\text{W}$, is the total exergy the cycle destroys. That $74\,\text{W}$
is the budget the component audit now distributes. The headline number is
already a warning: a domestic refrigerator wastes about five-sixths of
the work it draws, a far worse second-law showing than the steam plant's,
and the reason is that small temperature lifts make the Carnot benchmark
cheap while real components stay irreversible.

\subsection{Where the work goes, component by component}

Each component destroys exergy at the rate $T_0\,\dot S_{\text{gen}}$.
We take them in cycle order, carrying the entropy generation of each.

\paragraph{Compressor.} A real compressor raises the refrigerant
pressure with an isentropic efficiency near $0.70$ for a small hermetic
machine, so a good share of its electrical input goes into entropy rather
than into the pressure rise. Of the $90\,\text{W}$ drawn, motor and
mechanical losses plus the non-isentropic compression generate entropy at
a rate that destroys roughly $\dot X_{\text{d,comp}} \approx 16\,\text{W}$
of work potential, the largest single destruction. The compressor leads
the ledger because it is where the work enters and where the largest
single irreversibility, the gap between real and isentropic compression,
sits.

\paragraph{Condenser.} The condenser rejects the sum of the load and the
work, $\dot Q_H = \dot Q_C + \dot W_{\text{in}} = 120 + 90 =
210\,\text{W}$, into the kitchen air. The refrigerant condenses at a
temperature above ambient, perhaps $T_{\text{cond}} \approx 318\,\si{\kelvin}$,
so heat crosses a gap of about $15\,\si{\kelvin}$ from the refrigerant to
the $T_0 = 303\,\si{\kelvin}$ air. The entropy generated is
\[
\dot S_{\text{gen,cond}} \approx \dot Q_H
\left(\frac{1}{T_0} - \frac{1}{T_{\text{cond}}}\right)
= 210\left(\frac{1}{303} - \frac{1}{318}\right)
= 210 \times 1.56\times10^{-4}
= 0.033\,\si{\watt\per\kelvin},
\]
so the exergy destroyed is $T_0\,\dot S_{\text{gen,cond}} = 303 \times
0.033 \approx 10\,\text{W}$, a moderate share carried by the
finite-$\Delta T$ heat rejection.

\paragraph{Expansion valve.} The refrigerant leaves the condenser as a
high-pressure liquid at state $3$ and is throttled to the low pressure at
state $4$. The valve is adiabatic and does no work, so its enthalpy is
unchanged and a first-law balance writes nothing across it. Its entropy
nonetheless climbs, because dropping the pressure of a liquid through a
restriction is a textbook irreversibility: the pressure energy that an
ideal expander would have recovered as work is dissipated into the
fluid instead. The entropy generated is the refrigerant's entropy rise
across the valve times the mass flow, and for this cycle it destroys
about $\dot X_{\text{d,valve}} \approx 13\,\text{W}$ of work potential.
The throttle is the component the energy balance is blind to and the
exergy balance ranks second, almost level with the compressor. The
rightward jump $3 \!\to\! 4$ in
figure~\ref{fig:vol04ch04:refrigeration-ts} is exactly this destruction
drawn on the diagram.

\paragraph{Evaporator.} The evaporator absorbs the $120\,\text{W}$ load
from the cold space at a refrigerant temperature below the cold-space
temperature, perhaps $T_{\text{evap}} \approx 261\,\si{\kelvin}$ against a
cold space at $268\,\si{\kelvin}$. Heat crossing this $7\,\si{\kelvin}$
gap generates entropy, and because the whole evaporator sits below the
dead state the exergy bookkeeping carries the cold-space sign convention:
the cooling delivered is itself negative-temperature exergy relative to
$T_0$. The evaporator's finite-$\Delta T$ absorption destroys about
$\dot X_{\text{d,evap}} \approx 10\,\text{W}$ of work potential.

\subsection{The ledger and the verdict}

Figure~\ref{fig:vol04ch04:refrigeration-exergy-bars} sets the four
destructions side by side as percentages of the compressor work in. They
sum, with the cold-space exergy actually delivered, to the work supplied:
\[
\underbrace{16\,\text{W}}_{\text{compressor}} +
\underbrace{10\,\text{W}}_{\text{condenser}} +
\underbrace{13\,\text{W}}_{\text{valve}} +
\underbrace{10\,\text{W}}_{\text{evaporator}} +
\underbrace{41\,\text{W}}_{\text{delivered, motor, misc}}
= 90\,\text{W},
\]
the $90\,\text{W}$ of purchased work distributed term by term across
destruction and delivery. The compressor and the valve together account
for the larger part of the avoidable loss. The design responses follow
the ranking and differ sharply from what the energy balance would
suggest. The energy balance, seeing only that all the work leaves as heat
at the condenser, would point a designer at the condenser; the exergy
balance points instead at the compressor (a more efficient motor and a
better-matched compression) and at the valve. The valve loss is the
instructive one: replacing the throttle with a small expander that
recovers the pressure energy as work, or with a two-stage expansion that
breaks the pressure drop into steps, recovers a slice of the
$13\,\text{W}$ the simple valve discards. Industry keeps the cheap valve
because the recovered work is small in absolute terms for a domestic
unit, but the same calculation on a large industrial chiller, where the
valve loss is kilowatts, is what justifies the expander there.

The audit's lesson repeats the steam plant's in mirror image. The first
law conserves energy across every component and so cannot say where the
work is wasted; the throttle, which conserves enthalpy exactly, is its
blind spot. The second law weights each flow by its distance from the
dead state, charges each component the exergy it destroys, and so ranks
the components by where redesign pays. The discipline is the same in both
audits: follow the exergy, not the energy.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-refrigeration-ts.tex}

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-refrigeration-exergy-bars.tex}

\begin{mastery}
\textbf{Why the throttle survives the expander on a household unit}. The
$13\,\text{W}$ the valve destroys is the work an ideal expander would
recover, but the expander is not free. A liquid-vapour expander for a
domestic refrigerant flow is a small, sealed, high-speed machine whose
manufacture, sealing, and bearing losses cost more, in money and in their
own irreversibility, than the $13\,\text{W}$ they save is worth at
household duty. The work recovered scales with the mass flow and the
pressure drop; both are small for a kitchen appliance and large for an
industrial ammonia chiller. The crossover is an exergoeconomic
calculation: the value of the recovered exergy over the machine's life
against the capital and the parasitic loss of the recovery device. On a
large chiller the recovered kilowatts pay for the expander in a season;
on a refrigerator they never do, which is why the throttle, the largest
avoidable thermodynamic loss in the cycle, is also the correct
engineering choice at that scale. The second-law audit names the loss;
the economic audit decides whether it is worth removing. The two together
are exergoeconomics, the working discipline that keeps a second-law audit
from recommending a redesign that costs more than it saves.
\end{mastery}

\section{A third end-to-end case study: a fuel cell against a heat engine}\pathtag{standard}

The two exergy audits compared real devices against their own reversible
benchmarks and found the loss inside the hardware. A different second-law
question asks not how a device falls short of its own ideal but whether the
ideal itself can be raised by choosing a different conversion route. Hydrogen
carries chemical energy; an engineer can release that energy as heat and run a
cycle, bounded by Carnot, or release it as electrical work in a fuel cell,
bounded by the Gibbs free energy of the reaction. The two ceilings are not the
same, and the gap between them is the clearest demonstration in the chapter that
the Carnot bound governs heat engines, not energy conversion in general. We
carry a single hydrogen fuel cell and a single hydrogen-fired heat engine
through a head-to-head second-law comparison, with numbers, and then audit the
real cell's losses against its own ideal \cite{text:cengel-boles}.

\subsection{The two routes and their ceilings}

The reaction is the same in both routes, $\text{H}_2 + \tfrac12\text{O}_2
\rightarrow \text{H}_2\text{O}$, with the liquid-water product taken as the
basis. Its standard enthalpy of reaction is $\Delta H^\circ =
-286\,\text{kJ}/\text{mol}$, the higher heating value of hydrogen, and its
standard Gibbs free energy of reaction is $\Delta G^\circ =
-237\,\text{kJ}/\text{mol}$ at $T = 298\,\si{\kelvin}$. The difference,
$T\Delta S^\circ = \Delta H^\circ - \Delta G^\circ = -49\,\text{kJ}/\text{mol}$,
is the reaction's entropy term: the reaction consumes one and a half moles of
gas and produces a mole of liquid, a large drop in entropy, so $\Delta S^\circ
\approx -164\,\text{J}/(\text{mol}\cdot\si{\kelvin})$ is negative.

The heat-engine route burns the hydrogen, releasing $|\Delta H^\circ| =
286\,\text{kJ}/\text{mol}$ as heat at a flame temperature, and feeds that heat
to a cycle. The best the cycle can do is the Carnot fraction $1 - T_C/T_H$. For
a hot side at $T_H = 1500\,\si{\kelvin}$ against a cold sink at $T_C =
320\,\si{\kelvin}$, the ceiling is $1 - 320/1500 = 0.79$, and a real combined
cycle reaches perhaps $0.60$ of the heat. The fuel-cell route converts the
reaction's Gibbs free energy straight to electrical work without the heat step,
so its thermodynamic ceiling is
\[
\eta_{\text{fuel cell, ideal}} = \frac{|\Delta G^\circ|}{|\Delta H^\circ|}
= \frac{237}{286} = 0.83
\]
at room temperature. The cell's ceiling sits above the real engine and above
even the Carnot ceiling for any hot side below about $750\,\si{\kelvin}$, the
crossover that figure~\ref{fig:vol04ch04:fuelcell-vs-carnot} draws. The cell
escapes the Carnot bound not by violating the second law but by never invoking
it: Carnot bounds the conversion of heat to work, and the cell converts chemical
free energy to electrical work directly, a process the second law bounds by
$\Delta G$, not by $1 - T_C/T_H$.

\subsection{The reversible cell voltage}

The cell delivers its work as charge driven through a potential. Each hydrogen
molecule releases two electrons, so a mole of reaction moves $n = 2$ moles of
electrons, a charge $nF$ with $F = 96{,}485\,\text{C}/\text{mol}$ the Faraday
constant. The reversible work per mole is $|\Delta G^\circ|$, and the reversible
cell voltage is that work per unit charge,
\[
E_{\text{rev}} = \frac{|\Delta G^\circ|}{nF}
= \frac{237{,}000}{2 \times 96{,}485}
= 1.23\,\text{V}.
\]
This is the open-circuit voltage a hydrogen cell holds when it draws no current
and generates no entropy. The thermoneutral voltage, the value at which the cell
would produce no waste heat at all, is the larger
$|\Delta H^\circ|/(nF) = 286{,}000/(2\times96{,}485) = 1.48\,\text{V}$; the gap
between $1.48\,\text{V}$ and $1.23\,\text{V}$ is the reversible heat
$T\Delta S^\circ$ the cell must reject even when it runs perfectly, the
electrochemical signature of the negative reaction entropy.

\subsection{Where the real cell loses its work}

A real cell delivers less than $1.23\,\text{V}$ the moment it draws current,
and the deficit is the irreversibility the second law charges.
Figure~\ref{fig:vol04ch04:fuelcell-polarization} is the polarization curve: the
operating voltage falls from the reversible value through three loss regimes as
the current rises. At an operating point of $0.7\,\text{V}$, common for a
low-temperature cell at useful current density, the voltage efficiency is
\[
\eta_V = \frac{V_{\text{operating}}}{E_{\text{rev}}}
= \frac{0.70}{1.23} = 0.57,
\]
so the cell delivers $57\%$ of its own reversible ceiling, and the overall
work-from-enthalpy efficiency is $\eta_V \times |\Delta G^\circ|/|\Delta H^\circ|
= 0.57 \times 0.83 = 0.47$. We split the lost voltage by mechanism. The
activation overpotential, the kinetic price of driving the oxygen-reduction
reaction at the cathode, takes roughly $0.30\,\text{V}$ at this current and is
the largest single loss. The ohmic overpotential, the resistance of the proton
exchange membrane and the contacts, takes about $0.18\,\text{V}$. The
concentration overpotential, the cost of supplying reactant to the electrode
fast enough, takes about $0.05\,\text{V}$ here and grows steeply toward the
limiting current. The three sum to the $0.53\,\text{V}$ gap between $1.23$ and
$0.70$.

Each lost volt is exergy destroyed. Per mole of reaction the cell destroys
\[
\dot X_{\text{d,cell}} = nF\,(E_{\text{rev}} - V_{\text{operating}})
= 2 \times 96{,}485 \times 0.53
= 102\,\text{kJ}/\text{mol},
\]
appearing as waste heat inside the cell on top of the reversible
$T\Delta S^\circ = 49\,\text{kJ}/\text{mol}$ the reaction sheds even when
reversible. A cell drawing $0.7\,\text{A}/\text{cm}^2$ over a $100\,\text{cm}^2$
active area passes $70\,\text{A}$, so at $0.7\,\text{V}$ it delivers
$49\,\text{W}$ of electrical power while dissipating the activation, ohmic, and
concentration losses as heat that the cooling system must carry away.

\subsection{The verdict}

The second-law efficiency of the cell, work delivered over work available, is
the same $0.47$ figure computed two ways: the cell turns $47\%$ of the
hydrogen's enthalpy into electrical work, against a real heat engine's $0.60
\times 0.79 \approx 0.47$ for a high-temperature combined cycle and far more
than a single-cycle engine on a modest hot side. The comparison settles the
design question by route, not by polishing one device: at low and moderate
temperatures the electrochemical route starts from a higher ceiling because it
sidesteps Carnot, and its losses, dominated by the cathode activation
overpotential, are a catalysis problem rather than a thermodynamic one. The
engine route wins only where the hot side can be pushed past the
$750\,\si{\kelvin}$ crossover and the cell's materials cannot follow. The
broader lesson generalises past hydrogen: when a fuel's chemical energy can be
released as work without first becoming heat, the Carnot bound does not apply,
and the relevant ceiling is the Gibbs free energy. The code in
\texttt{code/exergy\_balance.py} reproduces the voltage-loss split and the
route comparison of figure~\ref{fig:vol04ch04:fuelcell-vs-carnot}.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-fuelcell-vs-carnot.tex}

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-fuelcell-polarization.tex}

\begin{mastery}
\textbf{Why a fuel cell can exceed the Carnot factor without breaking the second
law}. The claim that a fuel cell beats Carnot reads, on first hearing, like the
overunity claims the failure section dismisses. It is not, and the distinction
is exactly the one the failure section's first diagnostic demands: redraw the
boundary and name the process. Carnot bounds a cyclic device that takes heat
from a hot reservoir and rejects heat to a cold one; its ceiling $1 - T_C/T_H$
is the price of routing energy through a thermal reservoir. A fuel cell never
does this. Its input is not heat from a reservoir but the chemical free energy
of a reaction, and the second-law ceiling on extracting work from a reaction at
constant temperature and pressure is the Gibbs free energy $|\Delta G|$, which
can be a larger fraction of $|\Delta H|$ than any achievable Carnot factor. The
cell still obeys the second law: it cannot deliver more than $|\Delta G|$, it
sheds the reversible heat $T|\Delta S|$, and every real loss adds to that heat.
What it escapes is not the second law but the Carnot \emph{form} of the second
law, which applies only to heat engines. The reaction with positive reaction
entropy, such as the high-temperature oxidation of carbon to carbon monoxide,
$\Delta S > 0$, makes the point sharper still: for it $|\Delta G| > |\Delta H|$,
so an ideal cell running that reaction could deliver more electrical work than
the reaction's enthalpy, drawing the balance as heat from the surroundings, and
it still does not violate the second law because the heat it absorbs comes with
its entropy and the reaction's entropy increase pays for it. The lesson is that
``efficiency above Carnot'' is a warning to check the boundary, not an automatic
refutation; for a direct energy conversion that never passes through a thermal
reservoir, the Carnot factor is the wrong bound to apply.
\end{mastery}

\section{An applied vignette: the Earth as a planetary heat engine}\pathtag{standard}

The three case studies audited machines an engineer builds. The same accounting
applied to the planet itself explains why life and weather are possible at all,
and it makes the photon-gas entropy of the worked examples do a job at the largest
scale in the book. We carry a single steady-state entropy balance over the whole
Earth and find that the planet runs on an entropy budget, not an energy one.

The Earth in steady state absorbs almost exactly the solar power it re-radiates.
The absorbed solar power, after reflection, is about $\dot Q = 1.2\times10^{17}\,
\text{W}$, and in the long-term average the planet radiates the same power back to
space as infrared, because any lasting imbalance would heat or cool the planet
without bound. A first-law energy balance therefore closes with nothing left over:
energy in equals energy out, and by that ledger the Earth is doing no work and
gaining nothing. The engine is invisible to the first law, exactly as the boiler
was invisible in the steam-plant audit.

The entropy balance tells the real story. Radiation carries entropy at the rate
$\tfrac43$ of its power divided by its temperature, the photon-gas result of the
worked examples. The incoming sunlight arrives from the Sun's $T_{\text{sun}} =
5800\,\si{\kelvin}$ photosphere, so it delivers entropy at the rate
\[
\dot S_{\text{in}} = \frac{4}{3}\,\frac{\dot Q}{T_{\text{sun}}}
= \frac{4}{3}\times\frac{1.2\times10^{17}}{5800}
= 2.8\times10^{13}\,\si{\watt\per\kelvin}.
\]
The outgoing infrared leaves at the effective radiating temperature of the planet,
$T_{\text{earth}} \approx 255\,\si{\kelvin}$, so it carries entropy away at
\[
\dot S_{\text{out}} = \frac{4}{3}\,\frac{\dot Q}{T_{\text{earth}}}
= \frac{4}{3}\times\frac{1.2\times10^{17}}{255}
= 6.3\times10^{14}\,\si{\watt\per\kelvin}.
\]
The planet sheds about twenty-two times more entropy than it takes in, and the
difference,
\[
\dot S_{\text{export}} = \dot S_{\text{out}} - \dot S_{\text{in}}
\approx 6.0\times10^{14}\,\si{\watt\per\kelvin},
\]
is the entropy the Earth exports to space every second.
Figure~\ref{fig:vol04ch04:earth-entropy-budget} draws the ledger as a bar chart:
a small entropy inflow, a large entropy outflow, and the wide gap between them
that is the planet's export.

That export is the resource everything on the surface runs on. Photosynthesis
builds low-entropy sugars from high-entropy carbon dioxide and water, and it can
do so only because it degrades a handful of high-energy visible photons into a
larger number of low-energy infrared photons, exporting the entropy difference.
Weather systems, ocean currents, and the whole hydrological cycle are a heat
engine running between the warm tropics and the cold poles and radiating the waste
to space, their mechanical work paid for out of the same entropy export. The
biosphere does not violate the second law by building order; it pays for its order
by riding the entropy stream from a hot Sun to a cold sky, the planetary version
of the cell exporting entropy to its surroundings from section~4.4. If the Sun
delivered its energy at the same $255\,\si{\kelvin}$ the Earth radiates, the two
entropy fluxes would match, the export would vanish, and no engine, living or
meteorological, could run, even though the energy delivered would be unchanged.
The quality of the energy, its low entropy, not its quantity, is what makes the
planet habitable, the same distinction the exergy audits drew for the steam plant
and the refrigerator, written across the whole Earth.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-earth-entropy-budget.tex}

\section{An applied vignette: the vortex tube, a demon made of plumbing}\pathtag{standard}

A device that takes one stream of room-temperature compressed air and hands back
a cold stream and a hot stream, with no moving parts and no electricity, looks
on its face like Maxwell's demon built out of pipe fittings. The Ranque-Hilsch
vortex tube is exactly this device, and it is sold by the thousand for spot
cooling of machine tools and electronics. Its second-law accounting settles what
the demon argument only gestures at: the sorting is real, the temperature split
is real, and the second law is satisfied with room to spare, because the sorting
is paid for out of the entropy the supply air generates when its pressure
collapses.

Figure~\ref{fig:vol04ch04:vortex-tube} shows the layout. Compressed air enters
tangentially near one end and spins down a straight tube at high angular speed.
A throttling cone at the far end lets only the outer, slower-swirling layers
escape, and they leave hot; the inner core, forced to reverse and travel back up
the axis, leaves cold through an orifice at the near end. The energy split
follows from a first-law balance across the tube. If a fraction $\mu$ of the
inlet mass leaves cold and the rest leaves hot, and the tube exchanges no work
or heat with its surroundings, enthalpy is conserved:
\[
\dot m\, h_{\text{in}} = \mu \dot m\, h_{\text{cold}} + (1-\mu)\dot m\,
h_{\text{hot}},
\]
so the cold stream's enthalpy deficit is balanced by the hot stream's surplus.
For a supply at $T_{\text{in}} = 293\,\si{\kelvin}$ and $7\,\text{bar}$, a
typical tube tuned to a cold fraction $\mu = 0.3$ delivers cold air near
$253\,\si{\kelvin}$ ($-20\,\degC$) and hot air near $310\,\si{\kelvin}$
($37\,\degC$). The first law is content: the cold air's $40\,\si{\kelvin}$
deficit, weighted by its small mass fraction, matches the hot air's
$17\,\si{\kelvin}$ surplus, weighted by the larger fraction.

The second law is where the device stops looking like a demon. The demon of
section~4.5 sorts molecules without a cost until the memory bill is entered; the
vortex tube pays its bill up front, in the pressure drop of the supply air. The
inlet air at $7\,\text{bar}$ leaves at roughly atmospheric pressure, and that
expansion, spread across the turbulent swirl and the throttle, generates a large
entropy quite apart from the temperature sorting. We compare the two. The
entropy the sorting would have to erase, were the tube a demon, is of order the
mixing entropy of recombining the hot and cold streams, a few
$\si{\joule\per\kelvin}$ per mole. The entropy the pressure drop actually
generates is the isothermal-expansion entropy $R\ln(p_{\text{in}}/p_{\text{out}})
= 8.314\ln 7 = 16\,\si{\joule\per\kelvin}$ per mole, several times larger. The
tube generates far more entropy destroying the supply pressure than it removes
by sorting temperature, so the total entropy rises and the second law holds
comfortably. The vortex tube is a demon that pays retail: it buys its
temperature split with a pressure drop whose irreversibility dwarfs the split.

The exergy verdict is harsh and explains the device's niche. The compressed air
carries substantial exergy, all of which must first be manufactured by a
compressor running at its own second-law efficiency, and the vortex tube then
destroys the larger part of that exergy in the throttling swirl before delivering
a modest cold stream. Its second-law efficiency as a cooler is a few percent,
far below the vapour-compression refrigerator of the earlier case study. No
engineer would cool a building this way. The vortex tube survives for the same
reason the household throttle survives the expander: it trades thermodynamic
efficiency for mechanical simplicity, and where compressed air is already on hand
and reliability matters more than running cost, a fitting with no moving parts
that never wears out and needs no power beyond the air line is worth its
wastefulness. The demon that looked like a free lunch is a paid one, and the
receipt is written in the pressure gauge.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-vortex-tube.tex}

\section{A second applied vignette: the absorption refrigerator, cooling driven by heat}\pathtag{standard}

The vapour-compression refrigerator of the case study spent electrical work to
pump heat uphill. An absorption refrigerator does the same job with almost no
work at all, driven instead by a heat source: a gas flame, a district-heat
supply, or the waste heat of an engine. A refrigerator with no compressor and no
moving parts beyond a small pump, running on heat, is the kind of claim the
failure section teaches suspicion of, so it repays a careful second-law reading.
The device is honest; it is a heat engine and a heat pump fused into one, and the
second law bounds it by a product of two Carnot factors.

The working idea replaces the compressor's mechanical pressure rise with a
thermochemical one. A refrigerant, ammonia, is absorbed into a solvent, water,
at low pressure and temperature in an absorber; the ammonia-rich solution is
pumped, as a liquid, to high pressure at negligible work cost, because pumping a
liquid takes far less work than compressing a vapour; a generator then boils the
ammonia back out of solution using the heat source, delivering high-pressure
ammonia vapour to the condenser. From the condenser onward the cycle is an
ordinary vapour-compression loop: condense, throttle, evaporate in the cold
space, and return the low-pressure ammonia to the absorber to begin again. The
generator's heat has done the compressor's job of raising the refrigerant
pressure, at the price of a heat input rather than a work input.

The thermodynamic ceiling follows from treating the machine as two Carnot
devices in series. The heat source at $T_s$ drives a reversible engine rejecting
to the ambient $T_0$, which produces work at the Carnot efficiency $1 -
T_0/T_s$; that work then drives a reversible refrigerator lifting heat from the
cold space $T_C$ against the ambient, with coefficient of performance
$T_C/(T_0 - T_C)$. The ideal absorption coefficient of performance, cooling
delivered per unit heat supplied, is the product,
\[
\mathrm{COP}_{\text{abs}} = \left(1 - \frac{T_0}{T_s}\right)
\frac{T_C}{T_0 - T_C}.
\]
Take a gas-fired unit with a generator at $T_s = 450\,\si{\kelvin}$, an ambient
$T_0 = 300\,\si{\kelvin}$, and a cold space at $T_C = 275\,\si{\kelvin}$. The two
factors are $1 - 300/450 = 0.333$ and $275/(300 - 275) = 11.0$, so the ideal
coefficient of performance is
\[
\mathrm{COP}_{\text{abs}} = 0.333 \times 11.0 = 3.7.
\]
A real absorption unit reaches perhaps a fifth of this, an actual coefficient of
performance near $0.7$, well below a compression unit's $2$ to $4$ on
electrical work. The comparison is not apples to apples, and that is the point:
the compression unit is charged for high-quality electrical work, while the
absorption unit is charged for low-quality heat that might otherwise be dumped.
An absorption chiller running on the exhaust of a gas turbine, or on solar-heated
water, or on the waste heat of a data center, converts an exergy stream that
would otherwise be destroyed at the stack into cooling, and its poor coefficient
of performance on a first-law basis hides a strong second-law case: it is
spending exergy that was already paid for and about to be lost.

The failure-section diagnostic confirms the device is no perpetual machine.
Redraw the boundary: there are three heat reservoirs, not one, the hot generator,
the ambient, and the cold space, so the Kelvin-Planck prohibition on a
single-reservoir engine does not apply. Recount the flows: heat in at $T_s$,
heat out at $T_0$ from both the absorber and the condenser, and heat in at $T_C$
from the cold space, with a tiny pump work closing the balance. Recompute the
bound: the coefficient of performance cannot exceed the two-Carnot product above,
and no real unit approaches it. The absorption refrigerator is a legitimate
three-reservoir machine, the thermodynamic sibling of the vortex tube in that
both trade efficiency for the absence of a compressor, and the honest reading of
its low coefficient of performance is that it is a machine for spending cheap
exergy, not for making cold from nothing.

\section{A third applied vignette: the thermoacoustic refrigerator, cooling with sound}\pathtag{standard}

The absorption unit replaced the compressor with a chemistry; a thermoacoustic
refrigerator replaces it with a loudspeaker. A machine that pumps heat uphill using
nothing but a standing sound wave in a tube of pressurized gas, with no piston, no
refrigerant charge, and no moving part beyond the driver, is the kind of device the
failure section teaches suspicion of, so it repays a second-law reading. The device
is honest: it is a vapour-compression cycle with the compression stroke carried by
the acoustic pressure oscillation rather than by a mechanical piston, and the second
law bounds it by the same Carnot ceiling that bounds any refrigerator working
between the two temperatures.

Figure~\ref{fig:vol04ch04:thermoacoustic} shows the layout. A driver at one end of
a resonator sets up a half-wavelength standing wave in the gas, so the pressure
oscillates in step everywhere along the tube while the gas parcels slide back and
forth. A porous stack of plates sits between a hot heat exchanger and a cold one. A
gas parcel in the stack rides the standing wave through a small thermodynamic cycle
in each acoustic period: it is compressed and warmed as the pressure rises, it
touches the stack wall and gives up heat there while it is hot, it expands and cools
as the pressure falls, and it draws heat back from the wall a short distance away
while it is cold. Each parcel therefore carries a small quantity of heat one
parcel-step up the temperature gradient, and the chain of parcels along the stack is
a bucket brigade that moves heat from the cold exchanger, absorbing the load $Q_C$,
to the hot exchanger, rejecting $Q_H$ to ambient. The acoustic work the driver
supplies is the work input of the cycle, playing exactly the compressor's role.

The thermodynamic reading follows the vapour-compression audit of the earlier case
study point for point. The reversible ceiling is the Carnot heating-versus-cooling
coefficient of performance between the exchanger temperatures, and the real device
falls short of it by the entropy its irreversibilities generate: the finite
temperature difference between each gas parcel and the stack wall it exchanges heat
with, the viscous dissipation of the oscillating flow against the stack plates, and
the acoustic losses in the resonator. Take a unit lifting a $20\,\text{W}$ load from
a cold exchanger at $T_C = 250\,\si{\kelvin}$ to a hot exchanger at $T_H =
300\,\si{\kelvin}$. The Carnot cooling coefficient of performance is $T_C/(T_H -
T_C) = 250/50 = 5.0$, so the reversible acoustic work is $20/5.0 = 4\,\text{W}$; a
real thermoacoustic cooler reaches perhaps a third of the Carnot value, needing on
the order of $12\,\text{W}$ of acoustic power and destroying the difference in the
stack's finite-gap heat exchange and the viscous boundary layer. The imperfect
thermal contact that the stack needs to work, the parcel must exchange heat with the
wall, but not so fast that it equilibrates and stops shuttling, is the same
finite-gap irreversibility the counterflow-exchanger example charged, now doing
useful work rather than merely wasting it.

The failure-section diagnostic confirms the device is no free lunch. Redraw the
boundary: there are two heat reservoirs, the cold space and the ambient, plus the
acoustic work input, so this is an ordinary work-driven refrigerator, not a
single-reservoir engine. Recount the flows: work in from the driver, heat in at
$T_C$ from the load, heat out at $T_H$ to ambient, with $Q_H = Q_C + W$. Recompute
the bound: the coefficient of performance cannot exceed the Carnot value, and no
real unit approaches it. The thermoacoustic refrigerator earns its niche the way
the vortex tube and the absorption unit earn theirs, by trading efficiency for the
absence of a sliding seal: with no piston to wear and no refrigerant to leak, it
runs for years untended, and it can use an inert gas such as helium in place of a
halocarbon refrigerant, which is why it is built for cooling where reliability and
the working fluid matter more than the electricity bill, in spacecraft, in
liquefied-gas handling, and in cryogenic instruments. The sound that cools is paid
for in the same coin as every other refrigerator's work, and the receipt is the
acoustic power the driver draws.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-thermoacoustic.tex}

\section{The arrow of time and the second law as observation}\pathtag{mastery}

The mechanical laws (Newton's, Maxwell's, Einstein's, Schr\"odinger's)
are time-reversal invariant: if a film of a microscopic process is
played in reverse, the reversed process also satisfies the laws. Yet
macroscopic processes have an obvious temporal asymmetry: cream
disperses into coffee, never the reverse; broken eggs do not
spontaneously reassemble; a dropped wine glass shatters but does not
unshatter. The asymmetry has its origin in the statistical
interpretation of section 4.1 plus the initial conditions of the
observable universe.

The reasoning, due to Boltzmann and made precise across the twentieth
century, goes as follows. The microscopic equations of motion are
time-reversal invariant; any forward trajectory has a reverse
trajectory that also satisfies the equations. The asymmetry that we
observe must therefore come from the relative abundance of the two
classes of trajectories given the constraints we impose.

The constraint that produces the asymmetry is the initial condition
of the universe. The Big Bang began with the matter and radiation in
a state of very low entropy (very high uniformity, in a particular
sense), and the subsequent evolution has been a long approach toward
higher entropy. The thermodynamic arrow of time (the direction in
which entropy increases) coincides with the cosmological arrow
because both are determined by the same initial condition.

The reasoning is the working answer to ``why does the second law
hold?'' The law itself is an observation about the abundance of
microstates corresponding to different macrostates; the temporal
direction in which the abundance is realised is set by which end of
time has the lower entropy. Our universe began with low entropy, and
we live in its high-entropy expansion. The second law's
near-absoluteness is the practical consequence of the enormous
number of microstates involved.

Loschmidt's objection sharpens why the initial condition, and not the dynamics,
carries the arrow. Josef Loschmidt put the challenge to Boltzmann directly: since
the equations of motion are time-reversal invariant, for every trajectory along
which entropy rises there is a reversed trajectory, obtained by flipping every
velocity, along which entropy falls, and the two are equally valid solutions of
the mechanics. Why then does nature pick the rising one? The answer is that
nature does not pick; the initial conditions do. The velocity-reversed state that
would run entropy backward is a perfectly legal microstate, but it is one of the
vanishingly rare microstates finely tuned so that its future is a return to lower
entropy. The overwhelming majority of microstates consistent with a given
macrostate lie on rising-entropy trajectories, so a system prepared without such
fine tuning, which is every system an experimenter can actually prepare, almost
surely evolves toward higher entropy. The project of this chapter asks the reader
to see the point directly: reversing all the velocities of an equilibrated
simulated gas does send it briefly back toward its ordered start, confirming
Loschmidt, and the return is fragile, because the least numerical perturbation
knocks the trajectory off the fine-tuned reversed path and back onto a rising one.
The arrow is a statement about the abundance of trajectories and the poverty of
the fine-tuned ones, anchored at the low-entropy end by the initial condition of
the universe, not a property the equations of motion carry on their own.

The mastery box below sketches the harder questions, which the
engineering treatment in this chapter does not resolve.

\begin{mastery}
The second law's statistical origin raises several deep questions
that lie outside this volume's working scope but mark the boundary
of where the engineering treatment ends.

\textbf{Question of recurrence}. Poincar\'e showed in 1890 that an
isolated bounded dynamical system, given enough time, will return
arbitrarily close to its initial conditions. For a macroscopic
system, the recurrence time is unimaginably long (estimates for a
mole of gas range from $10^{10^{20}}$ to $10^{10^{25}}$ seconds), so
recurrence is operationally irrelevant to engineering. The recurrence
theorem makes the second law a statement about overwhelming
likelihood, not absolute impossibility.

\textbf{Question of measurement}. The macrostate-microstate
distinction depends on what is being measured. A coarser-grained
description (fewer measured macroscopic variables) has more
microstates per macrostate and more entropy; a finer-grained
description has fewer. Entropy is therefore observer-dependent in a
way that energy is not. The dependence does not affect engineering
practice (the macroscopic variables are set by the engineering
application), but it is a serious philosophical point.

\textbf{Question of the past hypothesis}. The low-entropy initial
condition of the universe is, in current cosmological theory, a
postulate rather than a derived consequence. Why was the universe in
a low-entropy state $13.8$ billion years ago? Several proposals
(eternal inflation, cyclic cosmology, anthropic selection, the
Penrose-Hawking conformal cyclic cosmology) attempt to answer; none
is yet settled.

These are open questions in physics and philosophy. They do not
affect the engineering use of entropy: the law is operationally
exact at the scales the engineer cares about, and the postmortem of
``why is it exact'' is a separate inquiry.
\end{mastery}

\section{Failure: statistical fluctuations vs systematic violations}\pathtag{core}

The boundary between a statistical fluctuation (rare but consistent
with the laws) and a systematic violation (a sign of misconception,
fraud, or undiscovered physics) is set by the magnitudes involved and
the conditions under which the observation was made. The chapter's
failure section catalogues the diagnostic patterns.

\subsection{Brownian fluctuations and similar phenomena}

A small mechanical system in thermal contact with a reservoir at
temperature $T$ exhibits fluctuations whose magnitudes are set by
the thermal energy $k_B T$. A torsion balance has angular
fluctuations of root-mean-square magnitude $\sqrt{k_B T/\kappa}$,
where $\kappa$ is the torsional stiffness; for $\kappa \sim 10^{-12}\,
\text{N}\cdot\text{m}/\text{rad}$ (a sensitive instrument), the rms
fluctuation is $\sim 2 \times 10^{-6}\,\text{rad}$ at room
temperature. A laser-cooled atomic gas at $T \sim 10\,\mu\text{K}$ has
much smaller fluctuations; a calorimeter at high $T$ has larger.

Such fluctuations look like local violations of the second law:
spontaneous decreases in entropy of small subsystems. They are
\emph{not} violations. The second law applies to the average over
many independent samples and to the total entropy of the system
including the reservoir; the subsystem's fluctuation is balanced by
the reservoir's. The diagnostic is the size of the fluctuation
relative to $k_B T$: a fluctuation of order $k_B T$ in a microscopic
system is expected; a fluctuation of order $N k_B T$ for macroscopic
$N$ would be extraordinary.

\subsection{Apparent violations in poorly calibrated measurements}

Routine engineering practice produces apparent violations of the
second law when measurements are made carelessly. The recurring
patterns:

\begin{itemize}[leftmargin=*]
  \item A claimed energy balance that does not close because an input
    flow was not measured (electric heater not on the same meter as
    the device being characterised; ambient heat leak into a hot
    calorimeter not deducted; mechanical work input not recorded).
  \item A claimed coefficient of performance above the Carnot bound
    because the operating temperatures were misidentified (the
    ``reservoir'' was actually two different temperatures at different
    times, with the COP being computed against the wrong reference).
  \item A claimed efficiency above $100\,\%$ because the input was
    quoted on lower heating value and the output on higher heating
    value (the condensing-furnace case of chapter~3).
\end{itemize}

The diagnostic, in each case, is to redraw the boundary, recount the
flows, and verify the balance with a different measurement instrument.
A second-law-violating claim that holds up to a careful redo of the
measurement is either a discovery (unlikely in 2026 given the law's
two-century empirical foundation) or a fundamental misconception
about what is being measured.

\subsection{Fluctuation theorems and the second law in small systems}

In the past three decades, theoretical and experimental work has
developed \engterm{fluctuation theorems} (Jarzynski equality, Crooks
fluctuation theorem) that quantify the probability of observing
non-equilibrium trajectories whose entropy change is negative. The
theorems state that a process with $\Delta S_{\text{traj}} = -|s|$
occurs with probability $\exp(-|s|/k_B)$ relative to a process with
$\Delta S_{\text{traj}} = +|s|$. For macroscopic $|s|$, the ratio is
astronomically small and the second law in its macroscopic form is
recovered. For microscopic systems (single molecules, nanoscale
devices), the ratio is observable and the fluctuation theorems are
experimentally verified.

The fluctuation theorems make the second law a statistical statement
that is exact at the average level and probabilistic at the trajectory
level. The engineering implication is that a sufficiently small
system can briefly appear to violate the second law; the violation
is not a violation of the law, but a particular trajectory of a
correctly-formulated stochastic process. The engineering of small-
scale machines (molecular motors, nanoscale devices, quantum thermal
machines) takes the fluctuation theorems as its operational
framework. The script \texttt{code/jarzynski\_demo.py} verifies the
Jarzynski equality $\langle e^{-W/k_B T}\rangle = e^{-\Delta F/k_B T}$
numerically for a dragged overdamped particle in a harmonic trap: the
mean work exceeds the free-energy difference (dissipation), yet the
exponential average of the work recovers the free-energy difference
exactly, the trajectory-level statement of the second law.

\subsection{Systematic-violation claims: the diagnostic}

A claim that the second law has been violated systematically (not as
a fluctuation in a small system) should be subjected to a standard
sequence of checks:

\begin{enumerate}[leftmargin=*]
  \item Redraw the boundary. Where are the inputs and outputs? Is
    there a hidden input (battery, electrical supply, chemical fuel,
    radioactive source)?
  \item Recount the flows. What is the actual mass, energy, and
    entropy passing each boundary?
  \item Recompute the bounds. What is the Carnot, the Kelvin-Planck,
    or the Clausius bound for the claimed process? Does the claim
    exceed the bound by a small amount (suggesting measurement error)
    or by an enormous amount (suggesting misconception)?
  \item Replicate. Is the result reproducible with the same protocol?
    By an independent measurement team with the same instruments?
    With different instruments?
\end{enumerate}

A claim that survives the four checks would be a discovery worthy of
the most careful subsequent scrutiny. The historical record contains
no such surviving claim across the two centuries since Carnot's
treatise; the failure section's working position is that the next
such claim will, like all its predecessors, dissolve under the four
checks.

\begin{principle}[Statistical asymmetry, not absolute prohibition]
The second law as a microscopic statement is statistical: a
macrostate with $W_2$ microstates is more likely than a macrostate
with $W_1$ microstates by the ratio $W_2/W_1$. For macroscopic $N$,
the ratio is exponentially enormous and the law is operationally
absolute. For microscopic $N$, the ratio is observable and the law
manifests as a fluctuation theorem. Engineering practice operates at
macroscopic $N$, so the law is exact within any reasonable
measurement; engineering of small-scale systems requires the
fluctuation-theorem framework, which is the same law expressed for
small-$N$ trajectories.
\end{principle}

\begin{estimation}
\textbf{Estimate the probability that all the air molecules in a
$1\,\text{m}^3$ room spontaneously occupy half the room.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The room contains $N \approx pV/(k_B T) = 10^5 \times 1/(1.38 \times
10^{-23} \times 300) \approx 2.4 \times 10^{25}$ molecules. Each
molecule, in random distribution, is equally likely to be in either
half of the room, so the probability that all $N$ are in one
specific half is $2^{-N}$.

Taking the logarithm: $\log_{10}(P) = -N \log_{10} 2 \approx -2.4
\times 10^{25} \times 0.301 \approx -7.2 \times 10^{24}$. The
probability is $10^{-(7 \times 10^{24})}$, a number so small it
requires a power tower to write.

For comparison, the age of the universe in Planck times is roughly
$10^{61}$. The number $10^{7 \times 10^{24}}$ exceeds the age of the
universe by $10^{7 \times 10^{24} - 61}$, which is still effectively
$10^{7 \times 10^{24}}$. The spontaneous-segregation event has
expected wait time of order $10^{7 \times 10^{24}}$ ages of the
universe.

The postmortem: the probability is so small that any actual
observation of such an event would force the immediate hypothesis
of measurement error or hidden input, not the hypothesis of a rare
fluctuation. The estimate makes concrete the gap between
``statistically allowed'' and ``effectively impossible'': the gap is
$10^{7 \times 10^{24}}$.
\end{estimation}

\section{Worked examples}\pathtag{standard}

The microscopic and free-energy machinery of the chapter resolves into a
handful of recurring calculations: how much entropy a real heat exchange
generates, which way a reaction runs and at what temperature it turns
around, and how the area of a cycle on a temperature-entropy diagram
delivers the work. We carry four of these through with numbers.

\subsection{Entropy generation in heat exchange across a finite gap}

A counterflow heat exchanger transfers $\dot Q = 50\,\text{kW}$ from a
hot oil stream entering at $T_H = 400\,\si{\kelvin}$ to a water stream
entering at $T_C = 320\,\si{\kelvin}$. We treat the two streams, to first
order, as large reservoirs at their mean temperatures and ask how much
entropy the exchange generates and how much exergy that destroys against
a $T_0 = 290\,\si{\kelvin}$ environment.

The hot stream sheds heat at $T_H$, losing entropy at the rate
$\dot Q/T_H = 50{,}000/400 = 125\,\si{\watt\per\kelvin}$. The cold stream
absorbs the same heat at $T_C$, gaining entropy at $\dot Q/T_C =
50{,}000/320 = 156\,\si{\watt\per\kelvin}$. The net entropy generated is
\[
\dot S_{\text{gen}} = \dot Q\left(\frac{1}{T_C} - \frac{1}{T_H}\right)
= 50{,}000\left(\frac{1}{320} - \frac{1}{400}\right)
= 50{,}000 \times 6.25\times10^{-4}
= 31\,\si{\watt\per\kelvin}.
\]
The exergy destroyed is $T_0\,\dot S_{\text{gen}} = 290 \times 31 =
9.1\,\text{kW}$: nearly a fifth of the heat transferred is stripped of
its capacity to do work, simply by crossing an $80\,\si{\kelvin}$ gap.
Figure~\ref{fig:vol04ch04:heatexchange-entropy} plots how this generation
grows with the gap; halving the approach to $40\,\si{\kelvin}$ roughly
halves the generation, at the cost of more heat-transfer area. The
calculation is the reason exergy-conscious design pushes streams toward
each other in temperature and prefers many small temperature steps to one
large one. A regenerator that breaks an $80\,\si{\kelvin}$ drop into eight
$10\,\si{\kelvin}$ steps generates roughly a tenth the entropy of the
single-step exchange, because the generation per step scales with the
square of the step.

The design tension this opens is the subject of
figure~\ref{fig:vol04ch04:egm-tradeoff}. Shrinking the approach
temperature reduces the heat-transfer irreversibility we just computed,
but the larger or faster channel that achieves it dissipates more pumping
work, a second source of entropy. The total entropy generation number
$N_s$ carries a minimum at an intermediate channel size, and designing to
that minimum, rather than to either term alone, is the
entropy-generation-minimisation method. For this exchanger, if the
pumping term rises linearly with throughput and the heat-transfer term
falls as its reciprocal, the optimum sits where the two are equal, and
the total generation there is twice the geometric mean of the two terms.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-heatexchange-entropy.tex}

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-egm-tradeoff.tex}

\subsection{Gibbs free energy and the spontaneity of a reaction}

The decomposition of calcium carbonate, $\text{CaCO}_3(\text{s})
\rightleftharpoons \text{CaO}(\text{s}) + \text{CO}_2(\text{g})$, is the
reaction that turns limestone into quicklime in a kiln. Its standard
enthalpy of reaction is $\Delta H^\circ = +178\,\text{kJ}/\text{mol}$
(strongly endothermic, heat must be supplied) and its standard entropy of
reaction is $\Delta S^\circ = +161\,\text{J}/(\text{mol}\cdot\si{\kelvin})$
(strongly positive, because a mole of gas appears where there was none).
We ask whether the reaction runs at room temperature, and at what
temperature it becomes spontaneous.

At $T = 298\,\si{\kelvin}$,
\[
\Delta G^\circ = \Delta H^\circ - T\Delta S^\circ
= 178{,}000 - 298 \times 161
= 178{,}000 - 47{,}978
= +130{,}000\,\text{J}/\text{mol}.
\]
The positive $\Delta G^\circ$ says the forward reaction does not run
spontaneously at room temperature; limestone is stable on the shelf. The
two terms pull in opposite directions: the endothermic $\Delta H^\circ$
opposes decomposition, while the entropy gain $-T\Delta S^\circ$ favours
it, and at room temperature the enthalpy wins. As temperature rises the
$-T\Delta S^\circ$ term grows in magnitude and eventually overturns the
balance. The crossover, where $\Delta G^\circ = 0$, is
\[
T^\ast = \frac{\Delta H^\circ}{\Delta S^\circ}
= \frac{178{,}000}{161}
= 1106\,\si{\kelvin} \approx 833\,\degC.
\]
Above about $833\,\degC$ the kiln decomposes the limestone
spontaneously; below it, the reaction will not proceed. The figure of
$1106\,\si{\kelvin}$ is close to the temperature at which the equilibrium
$\text{CO}_2$ partial pressure reaches one atmosphere, which is the
operating heuristic a kiln engineer uses. The sign pattern, endothermic
and entropy-increasing, is the signature of a reaction switched on by
temperature, the mirror of the water-gas-shift reaction earlier in the
chapter, which was exothermic and entropy-decreasing and so switched
\emph{off} by temperature.

\subsection{Carnot-cycle work as a temperature-entropy area}

A Carnot engine takes in heat from a reservoir at $T_H = 800\,\si{\kelvin}$,
rejects heat to a reservoir at $T_C = 300\,\si{\kelvin}$, and operates
with an entropy swing of $\Delta S = 1.5\,\si{\kilo\joule\per\kelvin}$
across the isothermal heat-addition stroke. We read its heat flows, work,
and efficiency directly off the temperature-entropy diagram of
figure~\ref{fig:vol04ch04:ts-diagram-carnot}, where the cycle is a
rectangle.

The heat absorbed along the top isotherm is the area under that edge,
\[
Q_H = T_H\,\Delta S = 800 \times 1.5 = 1200\,\text{kJ}.
\]
The heat rejected along the bottom isotherm is the area under it,
\[
Q_C = T_C\,\Delta S = 300 \times 1.5 = 450\,\text{kJ}.
\]
The net work is the enclosed area of the rectangle, the difference of the
two,
\[
W = (T_H - T_C)\,\Delta S = (800 - 300)\times 1.5 = 750\,\text{kJ},
\]
and the thermal efficiency is the ratio of the enclosed area to the area
under the top edge,
\[
\eta = \frac{W}{Q_H} = \frac{750}{1200} = 0.625
= 1 - \frac{T_C}{T_H} = 1 - \frac{300}{800} = 0.625,
\]
the two routes agreeing as they must. The temperature-entropy diagram
makes the second law geometric: the work is an area, the efficiency is a
ratio of areas, and a real cycle, which is not a clean rectangle because
its strokes generate entropy, encloses a smaller area than the Carnot
bound by exactly the $T_0\,\dot S_{\text{gen}}$ the Gouy-Stodola theorem
charges. Reading a real cycle's $T$-$S$ trace against its Carnot rectangle
is the graphical form of the exergy audit of the previous section.

\subsection{Entropy of mixing for an ideal solution}

Two liquids that form an ideal solution mix with no enthalpy change; the
free-energy driving force for mixing is entirely entropic. We compute the
entropy of mixing when $0.4\,\text{mol}$ of one component and
$0.6\,\text{mol}$ of a second are combined into one ideal solution at
constant temperature, and find the composition of maximum mixing entropy.

For an ideal solution the molar entropy of mixing per mole of solution is
\[
\Delta S_{\text{mix}} = -R\sum_i x_i \ln x_i,
\]
the same form as the ideal-gas mixing entropy, because an ideal solution
shares the ideal gas's assumption that the species do not interact
preferentially. With mole fractions $x_1 = 0.4$ and $x_2 = 0.6$,
\[
\Delta S_{\text{mix}} = -R\,(0.4\ln 0.4 + 0.6\ln 0.6)
= -8.314\,(0.4\times(-0.916) + 0.6\times(-0.511))
= -8.314\times(-0.673)
= 5.6\,\si{\joule\per\kelvin}
\]
per mole of solution. The total for one mole of mixture is therefore
$5.6\,\si{\joule\per\kelvin}$, positive, as mixing always is for an ideal
solution. The free-energy change is $\Delta G_{\text{mix}} = -T\Delta
S_{\text{mix}}$, which is negative at every temperature, so an ideal
solution always mixes spontaneously and never separates. Differentiating
$-R(x\ln x + (1-x)\ln(1-x))$ and setting the derivative to zero gives the
maximum at $x = 1/2$, the equimolar composition, where each mole of
solution carries $R\ln 2 = 5.76\,\si{\joule\per\kelvin}$ of mixing
entropy. The curve is the same combinatorial peak that
figure~\ref{fig:vol04ch04:entropy-of-mixing} draws for gases: zero at the
pure endpoints, maximal at the equimolar centre, and symmetric between
them. Real solutions depart from this ideal curve through their enthalpy
of mixing, which can make mixing exothermic (favouring solution) or
endothermic (and, if large enough, driving the components to separate
into two phases), but the entropic term always pushes toward mixing.

\subsection{Entropy of the universe for heat across a finite gap}

A direct way to see the second law generate entropy is to put a fixed
quantity of heat across a fixed temperature gap between two reservoirs and
add up the entropy on both sides. We pass $Q = 1000\,\text{J}$ from a hot
reservoir held at $T_H = 500\,\si{\kelvin}$ to a cold reservoir held at
$T_C = 300\,\si{\kelvin}$, both large enough that their temperatures do not
move, and compute the entropy change of each and of the two together.

The hot reservoir loses heat at its fixed temperature, so its entropy
falls by
\[
\Delta S_H = -\frac{Q}{T_H} = -\frac{1000}{500} = -2.0\,\si{\joule\per\kelvin}.
\]
The cold reservoir gains the same heat at its fixed temperature, so its
entropy rises by
\[
\Delta S_C = +\frac{Q}{T_C} = +\frac{1000}{300} = +3.33\,\si{\joule\per\kelvin}.
\]
The entropy of the universe, the sum of the two, is
\[
\Delta S_{\text{univ}} = \Delta S_H + \Delta S_C
= -2.0 + 3.33 = +1.33\,\si{\joule\per\kelvin} > 0,
\]
positive, as the second law requires for a spontaneous process. The same
$1000\,\text{J}$ left the hot side carrying $2.0\,\si{\joule\per\kelvin}$
and arrived on the cold side carrying $3.33\,\si{\joule\per\kelvin}$,
because entropy carried by heat is $Q/T$ and the cold side's lower $T$
divides the same $Q$ into more entropy. The gap manufactured
$1.33\,\si{\joule\per\kelvin}$ from nothing, and against a $T_0 =
300\,\si{\kelvin}$ environment that destroys $T_0\,\Delta S_{\text{univ}}
= 300 \times 1.33 = 400\,\text{J}$ of work potential, two-fifths of the
heat transferred, lost simply by letting it slide down the gap rather than
routing it through an engine. The entropy generated scales with the gap:
were the gap shrunk to $T_H = 320\,\si{\kelvin}$ against $T_C =
300\,\si{\kelvin}$, the same $1000\,\text{J}$ would generate only
$1000(1/300 - 1/320) = 0.21\,\si{\joule\per\kelvin}$, a sixth as much. The
lesson repeats the heat-exchanger example: never drop heat across a wider
gap than the duty demands, because the gap is where the work potential
dies.

\subsection{Available work from a finite hot body cooling to ambient}

A reservoir holds its temperature; a finite hot body does not. We have a
$50\,\text{kg}$ steel forging at $T_i = 800\,\si{\kelvin}$ that will cool
to the ambient $T_0 = 300\,\si{\kelvin}$, with specific heat $c =
0.46\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$, and ask the maximum work
recoverable as it cools. The body is not a reservoir, so the Carnot factor
slides as it cools, and the available work is an integral, not a product.

The heat the body sheds in cooling by $dT$ is $dQ = mc\,(-dT)$ (negative,
because the body loses heat). An ideal engine drawing this heat at the
body's current temperature $T$ and rejecting to $T_0$ converts the
fraction $1 - T_0/T$ to work. The maximum work is therefore
\[
W_{\max} = \int_{T_0}^{T_i} \left(1 - \frac{T_0}{T}\right) mc\,dT
= mc\left[(T_i - T_0) - T_0 \ln\frac{T_i}{T_0}\right].
\]
With $mc = 50 \times 460 = 23{,}000\,\text{J}/\si{\kelvin}$,
\[
W_{\max} = 23{,}000\left[(800 - 300) - 300 \ln\frac{800}{300}\right]
= 23{,}000\left[500 - 300 \times 0.981\right]
= 23{,}000 \times 206 = 4.7\,\text{MJ}.
\]
The total heat the body sheds in reaching ambient is $Q = mc(T_i - T_0) =
23{,}000 \times 500 = 11.5\,\text{MJ}$, so the recoverable work is
$4.7/11.5 = 41\%$ of the heat. A reservoir fixed at $800\,\si{\kelvin}$
would yield the Carnot fraction $1 - 300/800 = 0.625$ of every joule; the
finite body yields only $41\%$, because each later joule leaves at a lower
temperature and so converts at a smaller Carnot factor.
Figure~\ref{fig:vol04ch04:finite-reservoir-exergy} shows the factor
sliding down the cooling curve, with the recoverable work the area under
it. The result is the exergy of a finite thermal mass, the number that
sets what a waste-heat-recovery engine can in principle take from a hot
slug of metal, exhaust gas, or quenching bath.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-finite-reservoir-exergy.tex}

\subsection{Statistical entropy of a freely jointed polymer chain}

The microscopic side of the chapter resolves into a count of states, and
a polymer chain makes that count tangible. A freely jointed chain of $N$
rigid links, each of length $b$, models a flexible polymer as a random
walk: each link points in a direction independent of its neighbours. We
compute the entropy as a function of the chain's end-to-end extension and
recover the entropic spring, the origin of rubber elasticity.

A chain stretched to end-to-end vector $\mathbf{R}$ has fewer accessible
configurations than a coiled chain, because most arrangements of the links
leave the ends near each other and only a few stretch them apart. For a
long chain the number of configurations reaching extension $R$ follows the
Gaussian count
\[
W(R) \propto \exp\!\left(-\frac{3R^2}{2N b^2}\right),
\]
the random-walk end-to-end distribution. The entropy is $S = k_B \ln W$,
so
\[
S(R) = S_0 - \frac{3 k_B R^2}{2 N b^2},
\]
a parabola falling from its maximum at $R = 0$. Stretching the chain
lowers its entropy, and the free energy $F = U - TS$ at fixed internal
energy (the links carry no energy, only orientation) is purely entropic,
$F(R) = F_0 + 3 k_B T R^2/(2 N b^2)$. The restoring force is the gradient
of the free energy,
\[
f = \frac{\partial F}{\partial R} = \frac{3 k_B T}{N b^2}\,R,
\]
a Hooke's-law spring whose stiffness $3 k_B T/(N b^2)$ rises with
temperature. A rubber band pulls back harder when warm, the opposite of a
metal spring, because its elasticity is the entropy fighting to recoil the
chain into its many coiled states rather than energy stored in stretched
bonds. The same count, applied to a hundred links with $b =
0.25\,\text{nm}$, gives a single-chain stiffness of order $3 \times 1.38
\times 10^{-23} \times 300/(100 \times (0.25\times10^{-9})^2) \approx 2
\times 10^{-3}\,\text{N}/\text{m}$, and a macroscopic rubber sheet's
modulus is this per-chain stiffness multiplied by the chain number
density, the bridge from the microstate count to a bulk mechanical
property.

\subsection{Counting microstates: the Einstein solid and $S = k_B \ln W$}

The formula $S = k_B \ln W$ is abstract until a count is carried out
explicitly. The Einstein solid, $N$ independent quantum harmonic oscillators
sharing $q$ identical quanta of energy, gives the cleanest closed-form count and
recovers a temperature, a heat capacity, and the third law from pure
combinatorics. We count the microstates, take the logarithm, and read off the
thermodynamics for a small crystal.

The number of ways to distribute $q$ indistinguishable quanta among $N$
distinguishable oscillators is the multiplicity
\[
W(N, q) = \binom{q + N - 1}{q}
= \frac{(q + N - 1)!}{q!\,(N-1)!},
\]
the standard stars-and-bars count: $q$ stars and $N-1$ bars in a row. Take a
crystal of $N = 3 \times 10^{4}$ oscillators (ten thousand atoms, three
directions each) holding $q = 3 \times 10^{4}$ quanta, one per oscillator on
average. The entropy is $S = k_B \ln W$, and for large $N$ and $q$ Stirling's
approximation turns the logarithm into
\[
\frac{S}{k_B} = \ln W \approx (q+N)\ln(q+N) - q\ln q - N\ln N.
\]
With $q = N$ this is $2N\ln(2N) - N\ln N - N\ln N = 2N\ln 2$, so
$S = 2N k_B \ln 2 = 2 \times 3\times10^4 \times 1.381\times10^{-23} \times
0.693 = 5.7\times10^{-19}\,\si{\joule\per\kelvin}$ for this small crystal,
matching the order $k_B N$ that the chapter's opening estimate gave for a
macroscopic count scaled down by Avogadro's number.

The temperature follows from $1/T = (\partial S/\partial E)_{N}$ with energy
$E = q\,\hbar\omega$, so $\partial S/\partial E = (1/\hbar\omega)\,\partial
S/\partial q$. Differentiating the Stirling form,
\[
\frac{\partial (S/k_B)}{\partial q} = \ln\frac{q+N}{q}
= \ln\left(1 + \frac{N}{q}\right),
\]
which gives
\[
\frac{1}{T} = \frac{k_B}{\hbar\omega}\ln\left(1 + \frac{N}{q}\right),
\qquad
q = \frac{N}{e^{\hbar\omega/k_B T} - 1}.
\]
The second form is the Bose-Einstein occupation of a single oscillator
multiplied by $N$, recovered here from a bare count of arrangements. As $T \to
0$ the quanta empty out, $q \to 0$, $W \to 1$, and $S \to 0$: the third law of
thermodynamics is the statement that a system with a unique ground state has one
accessible microstate at absolute zero, and the count delivers it. The
heat capacity $C = dE/dT$ taken from $E = N\hbar\omega/(e^{\hbar\omega/k_B T}-1)$
is the Einstein heat-capacity curve, $3Nk_B$ at high temperature (equipartition)
and falling exponentially toward zero at low temperature as the quanta freeze
out, the historical first explanation of why solids violate the
Dulong-Petit law in the cold. The whole of the solid's thermodynamics here comes
from one binomial coefficient and its logarithm.

\subsection{Entropy-generation minimisation of a counterflow heat exchanger}

The heat-exchange example earlier in this section computed the entropy a finite
temperature gap generates and noted that shrinking the gap costs pumping power.
We now make that tension quantitative and locate the optimum, the
entropy-generation-minimisation result that sets the best achievable
irreversibility of a heat exchanger. A counterflow gas exchanger carries two
contributions to entropy generation: the heat-transfer term, from passing heat
across the stream-to-stream temperature difference, and the fluid-friction term,
from pushing the gas through the channels against pressure drop. Bejan's
dimensionless entropy-generation number collects both.

For a gas stream of mass flow $\dot m$ and specific heat $c_p$ exchanging heat
$\dot Q$ across a small temperature difference $\Delta T$ around a mean
temperature $T$, the heat-transfer entropy generation is
\[
\dot S_{\text{gen},\Delta T} \approx \frac{\dot Q\,\Delta T}{T^2},
\]
the small-gap limit of $\dot Q(1/T_C - 1/T_H)$. The friction entropy generation
from a pressure drop $\Delta p$ pushed through at temperature $T$ is
\[
\dot S_{\text{gen},\Delta p} \approx \frac{\dot m\,R\,\Delta p}{p\,T}\cdot T
= \frac{\dot m R\,\Delta p}{p},
\]
for an ideal gas, the dissipated flow work divided by temperature. The heat
exchanger's geometry trades the two: a longer, narrower channel shrinks $\Delta
T$ (more area, closer thermal approach) but raises $\Delta p$ (more friction).
Writing both in terms of a single design variable, the duct's hydraulic
diameter or equivalently the flow Reynolds number $Re$, the heat-transfer term
falls as $Re^{-1}$ while the friction term rises as $Re^{2}$ or so, and the
total entropy generation
\[
N_s = \frac{\dot S_{\text{gen}}}{\dot S_{\text{gen,min}}}
= \frac{1}{2}\left(\frac{Re_{\text{opt}}}{Re} + \frac{Re}{Re_{\text{opt}}}\right)
\]
in the canonical scaling carries a minimum where the two contributions are
equal. Setting $d\dot S_{\text{gen}}/dRe = 0$ gives the optimum Reynolds
number, and at that point the total generation is twice the geometric mean of
the two terms, with $N_s = 1$ at the optimum by construction. For a concrete
case, take a stream where the heat-transfer term alone gives
$\dot S_{\text{gen},\Delta T} = 40/Re\,\si{\watt\per\kelvin}$ (in scaled units)
and the friction term gives $\dot S_{\text{gen},\Delta p} =
1\times10^{-5}\,Re\,\si{\watt\per\kelvin}$. The total
$\dot S_{\text{gen}} = 40/Re + 10^{-5}Re$ is minimised at
\[
Re_{\text{opt}} = \sqrt{\frac{40}{10^{-5}}} = \sqrt{4\times10^{6}} = 2000,
\]
where each term equals $40/2000 = 0.02\,\si{\watt\per\kelvin}$ and the total is
$0.04\,\si{\watt\per\kelvin}$, twice the geometric mean as promised. Running the
exchanger faster than $Re = 2000$ wastes the gain to friction; running it slower
wastes it to a wide thermal approach. The minimum is shallow near the optimum,
so a designer has room to trade a small entropy penalty for compactness or cost,
which figure~\ref{fig:vol04ch04:egm-tradeoff} shows as the flat valley of the
$N_s$ curve. The method generalises to any component with competing
irreversibilities; the working move is to write each contribution against one
design variable and minimise the sum, not either term alone.

\subsection{Verifying the Clausius inequality on a real cycle}

The Clausius inequality $\oint \delta Q/T \le 0$ is a bound, and a numerical
check on a real cycle shows the inequality strict and measures the entropy
generated. We take an internally reversible engine that exchanges heat with two
reservoirs but does so across finite temperature gaps, the standard model of an
endoreversible engine, and evaluate the cyclic integral.

The working fluid absorbs $Q_H = 1000\,\text{J}$ from a hot reservoir at
$T_H = 600\,\si{\kelvin}$ but, because the heat crosses a finite gap, the fluid
is at only $T_{H}' = 560\,\si{\kelvin}$ while it absorbs. It rejects $Q_C =
650\,\text{J}$ to a cold reservoir at $T_C = 300\,\si{\kelvin}$ while the fluid
sits at $T_{C}' = 320\,\si{\kelvin}$. The cyclic integral evaluated at the
\emph{fluid's} temperatures, where the heat actually crosses the boundary into
the fluid, is
\[
\oint \frac{\delta Q}{T} = \frac{Q_H}{T_H'} - \frac{Q_C}{T_C'}
= \frac{1000}{560} - \frac{650}{320}
= 1.786 - 2.031 = -0.245\,\si{\joule\per\kelvin} < 0,
\]
strictly negative, confirming the inequality for this irreversible cycle and
showing the working fluid itself returns to its state with $\oint dS = 0$ while
$\oint \delta Q/T < 0$, the difference being the entropy generated inside the
fluid by the irreversible heat admission. The total entropy generated per cycle
includes the reservoirs: the hot reservoir loses $Q_H/T_H = 1000/600 =
1.667\,\si{\joule\per\kelvin}$, the cold reservoir gains $Q_C/T_C = 650/300 =
2.167\,\si{\joule\per\kelvin}$, and the fluid returns to its start with no net
change, so
\[
\dot S_{\text{gen}} = -1.667 + 2.167 = +0.500\,\si{\joule\per\kelvin} > 0
\]
per cycle, positive as the second law requires. The cycle's work is $W = Q_H -
Q_C = 350\,\text{J}$ and its efficiency $\eta = 350/1000 = 0.35$, below the
Carnot ceiling $1 - 300/600 = 0.50$ for these reservoirs; the shortfall is the
exergy destroyed in the two finite gaps, $T_0\,\dot S_{\text{gen}} = 300 \times
0.5 = 150\,\text{J}$ against a $300\,\si{\kelvin}$ environment, exactly the gap
between the Carnot work $500\,\text{J}$ and the actual $350\,\text{J}$. The
Clausius inequality, the entropy balance, and the exergy accounting are three
readings of the same arithmetic, and a real cycle satisfies all three with the
same numbers.

\subsection{The general $c_p - c_v$ relation from a Maxwell identity}

The two heat capacities of a substance differ, and the gap is set entirely by the
substance's measurable thermal and mechanical response. We derive the general
relation $c_p - c_v = T V \alpha^2/\kappa_T$ from a Maxwell relation and evaluate
it for water, to show that the familiar $c_p - c_v = R$ of an ideal gas is one
limit of a result that holds for solids and liquids too.

Start from the entropy as a function of temperature and volume, $S(T,V)$, whose
differential is $dS = (\partial S/\partial T)_V\,dT + (\partial S/\partial
V)_T\,dV$. The first coefficient is $c_v/T$ by definition; the second is
$(\partial p/\partial T)_V$ by the third Maxwell relation. Compute the heat
capacity at constant pressure, $c_p = T(\partial S/\partial T)_p$, by dividing the
differential by $dT$ at fixed pressure,
\[
c_p = T\left(\pdv{S}{T}\right)_p
= c_v + T\left(\pdv{p}{T}\right)_V\left(\pdv{V}{T}\right)_p.
\]
The two derivatives on the right are written through the volume
expansivity $\alpha = (1/V)(\partial V/\partial T)_p$ and the isothermal
compressibility $\kappa_T = -(1/V)(\partial V/\partial p)_T$, using the triple
product rule $(\partial p/\partial T)_V = \alpha/\kappa_T$, to give
\[
c_p - c_v = \frac{T V \alpha^2}{\kappa_T}.
\]
The right side is non-negative for every stable substance, because $\kappa_T > 0$
for mechanical stability and $\alpha^2 \ge 0$, so $c_p \ge c_v$ always. For an
ideal gas $\alpha = 1/T$ and $\kappa_T = 1/p$, so $c_p - c_v = TV(1/T^2)p = pV/T =
R$ per mole, the schoolroom result. For liquid water at $25\,\degC$, with $\alpha
\approx 2.6\times10^{-4}\,/\si{\kelvin}$, $\kappa_T \approx
4.5\times10^{-10}\,\text{Pa}^{-1}$, and molar volume $V_m \approx
1.8\times10^{-5}\,\text{m}^3/\text{mol}$,
\[
c_p - c_v \approx \frac{298 \times 1.8\times10^{-5} \times (2.6\times10^{-4})^2}
{4.5\times10^{-10}}
\approx 0.8\,\si{\joule\per\kelvin}\,\text{per mole},
\]
a small fraction of water's $c_p \approx 75\,\si{\joule\per\kelvin}$ per mole. The
two heat capacities of a liquid or solid nearly coincide, which is why handbooks
quote a single value, while for a gas they differ by the full $R$. The general
relation says why: the gap scales with the expansivity squared, and a
nearly incompressible liquid expands little on heating.

\subsection{The Joule-Thomson coefficient and the inversion temperature}

A real gas forced through a throttle at constant enthalpy changes temperature, and
the sign of the change decides whether a gas can be liquefied by throttling. We
compute the Joule-Thomson coefficient from a Maxwell relation and find the
inversion temperature for a van der Waals gas.

The Joule-Thomson coefficient is the temperature change per unit pressure drop at
constant enthalpy, $\mu_{\text{JT}} = (\partial T/\partial p)_H$. The triple
product rule turns this into
\[
\mu_{\text{JT}} = -\frac{1}{c_p}\left(\pdv{H}{p}\right)_T,
\]
and the enthalpy's pressure dependence follows from $dH = T\,dS + V\,dp$ together
with the fourth Maxwell relation $(\partial S/\partial p)_T = -(\partial
V/\partial T)_p$, giving
\[
\left(\pdv{H}{p}\right)_T = V - T\left(\pdv{V}{T}\right)_p,
\qquad
\mu_{\text{JT}} = \frac{1}{c_p}\left[T\left(\pdv{V}{T}\right)_p - V\right].
\]
For an ideal gas $T(\partial V/\partial T)_p = V$, so $\mu_{\text{JT}} = 0$: an
ideal gas does not change temperature on throttling. A real gas does. To first
order a van der Waals gas has $\mu_{\text{JT}} \approx (1/c_p)(2a/RT - b)$, which
changes sign at the \engterm{inversion temperature}
\[
T_{\text{inv}} = \frac{2a}{Rb}.
\]
Below the inversion temperature $\mu_{\text{JT}} > 0$ and throttling cools the gas;
above it throttling warms the gas. For nitrogen, with $a = 0.137\,\text{Pa}\cdot
\text{m}^6/\text{mol}^2$ and $b = 3.87\times10^{-5}\,\text{m}^3/\text{mol}$, the
estimate gives $T_{\text{inv}} = 2(0.137)/(8.314 \times 3.87\times10^{-5}) \approx
850\,\si{\kelvin}$, comfortably above room temperature, so nitrogen cools on
throttling at ambient conditions and can be liquefied by repeated
throttle-and-cool stages. Hydrogen and helium have inversion temperatures below
room temperature ($\approx 200\,\si{\kelvin}$ and $\approx 40\,\si{\kelvin}$), so
they must be pre-cooled below inversion before throttling will liquefy them, the
reason early attempts to liquefy helium by simple throttling failed. The Maxwell
relation turns the sign of a gas-cooling process into a property of the
substance's equation of state.

\subsection{Absolute entropy from heat-capacity integration and the third law}

The third law fixes the absolute entropy at zero, and integrating the measured
heat capacity from absolute zero builds the entropy at any temperature. We compute
the standard molar entropy of a solid from its low-temperature Debye behaviour and
its tabulated heat capacity, and show how the third law makes the integral
convergent.

The absolute entropy at temperature $T$ is
\[
S(T) = \int_0^T \frac{c_p(T')}{T'}\,dT'
+ \sum_{\text{transitions}} \frac{\Delta H_{\text{trans}}}{T_{\text{trans}}},
\]
the heat-capacity integral plus a latent-heat term at each phase transition. The
integrand $c_p/T'$ would diverge at the lower limit if $c_p$ stayed finite, but
the third law forces $c_p \to 0$, and for a crystalline solid the Debye law
$c_p \approx a T'^3$ near absolute zero makes the integrand $a T'^2$, which
integrates to $a T^3/3 = c_p(T)/3$, a finite contribution. Take a solid whose
heat capacity follows $c_p = a T^3$ up to a matching temperature $T_1 =
15\,\si{\kelvin}$, with $a$ fixed so that $c_p(T_1) = 3.0\,\si{\joule\per\kelvin}$
per mole, giving $a = 3.0/15^3 = 8.9\times10^{-4}\,\text{J}/(\text{mol}\cdot
\text{K}^4)$. The entropy gathered from $0$ to $T_1$ is
\[
S(T_1) = \int_0^{T_1} \frac{a T'^3}{T'}\,dT' = \frac{a T_1^3}{3}
= \frac{c_p(T_1)}{3} = \frac{3.0}{3} = 1.0\,\si{\joule\per\kelvin}
\]
per mole, the closed-form benefit of the cubic law: the entropy below the matching
temperature is one third of the heat capacity there. Above $T_1$ the integral is
carried numerically from tabulated $c_p$ data, area under the $c_p/T$ curve, and
any melting or structural transition adds its $\Delta H_{\text{trans}}/
T_{\text{trans}}$. The procedure is the route by which every standard molar
entropy in the thermochemical tables is established, and the third law is what
makes the lower limit of the integral a definite zero rather than an unknown
constant. A residual entropy at absolute zero, as in carbon monoxide frozen with
its molecules pointing two ways, signals a ground state that is not unique, $W >
1$, and the count $S(0) = k_B \ln W$ recovers the measured residual exactly.

\subsection{The Landauer cost of erasing a hard drive}

The information-entropy preview gave the Landauer bound per bit; a complete drive
makes the number tangible and shows where the thermodynamic floor sits against the
energy a real erase actually costs. We compute the Landauer minimum heat for
clearing a modern hard drive and weigh it against the electrical energy the drive
draws to do the job.

Take a drive holding $C = 4\,\text{TB}$, which is $C = 4\times10^{12}\times 8 =
3.2\times10^{13}$ bits. Erasing one bit at room temperature must release at least
the Landauer heat $k_B T \ln 2 = 1.381\times10^{-23}\times 300\times 0.693 =
2.87\times10^{-21}\,\text{J}$. The thermodynamic floor for clearing the whole
drive is
\[
Q_{\min} = C\,k_B T \ln 2
= 3.2\times10^{13}\times 2.87\times10^{-21}
= 9.2\times10^{-8}\,\text{J},
\]
about $0.1\,\mu\text{J}$, the heat a hand releases in a fraction of a microsecond.
The floor is so far below any practical energy that it never constrains storage
design. A real overwrite erase reads and writes every sector: a drive drawing
$\sim 8\,\text{W}$ for the $\sim 6\,\text{hours}$ a full multi-pass wipe of a
$4\,\text{TB}$ disk takes consumes $8\times 6\times 3600 \approx
1.7\times10^{5}\,\text{J}$, more than twelve orders of magnitude above the
Landauer floor. Figure~\ref{fig:vol04ch04:landauer-erasure} sets the floor against
this and other per-bit budgets on a logarithmic axis; the floor is a thin line at
the far left of a chart whose real bars sit decades to its right.

The vignette carries a second lesson past the arithmetic. The Landauer cost is
charged for \emph{logically irreversible} operations, the ones that destroy the
record of their input, and erasure is the canonical case: collapsing a bit that
could have been $0$ or $1$ into a definite $0$ halves the memory's phase-space
volume, an entropy decrease $k_B \ln 2$ in the memory that the second law requires
be paid by at least $k_B T \ln 2$ of heat exported to the surroundings. A drive
that merely flips bits without forgetting their prior value, in principle, owes
nothing to Landauer; the cost is tied to the forgetting, not the writing. The
secure-erase problem is therefore a thermodynamic one only at a floor no engineer
will ever feel, and a practical one entirely about the read-write energy and the
time the wipe takes. The gap between the two, the twelve decades on the figure, is
the same headroom that the failure section's overunity claims never find a way to
close: real irreversibility, not the thermodynamic floor, sets the energy bill.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-landauer-erasure.tex}

\subsection{The entropy budget of melting ice in a warm drink}

A glass of water with ice in it equilibrates by melting the ice, and the entropy
the process generates is a direct, everyday instance of the second law operating
across a finite temperature gap. We compute the entropy change of the ice, of the
water, and of the two together, and show the total is positive, the spontaneous
direction.

Drop a $30\,\text{g}$ ice cube at $0\,\degC$ into $150\,\text{g}$ of water at
$25\,\degC$ in an insulated glass, and let it reach a common final temperature.
The latent heat of fusion is $L = 334\,\text{kJ}/\text{kg}$ and the specific heat
of liquid water is $c = 4.18\,\text{kJ}/(\text{kg}\cdot\si{\kelvin})$. First find
the final temperature from an energy balance: the heat to melt the ice and then
warm the meltwater equals the heat the warm water sheds,
\[
m_i L + m_i c (T_f - 273.15) = m_w c (298.15 - T_f).
\]
With $m_i = 0.030\,\text{kg}$ and $m_w = 0.150\,\text{kg}$, the melting demand is
$m_i L = 0.030\times 334{,}000 = 10{,}020\,\text{J}$, and solving the balance
gives $T_f \approx 280.7\,\si{\kelvin}$ ($\approx 7.5\,\degC$). The entropy
changes follow term by term. The ice melts at the fixed $273.15\,\si{\kelvin}$,
gaining
\[
\Delta S_{\text{melt}} = \frac{m_i L}{273.15}
= \frac{10{,}020}{273.15} = 36.7\,\si{\joule\per\kelvin}.
\]
The meltwater then warms from $273.15$ to $T_f$, gaining
\[
\Delta S_{\text{warm}} = m_i c \ln\frac{T_f}{273.15}
= 0.030\times 4180 \times \ln\frac{280.7}{273.15}
= 125.4 \times 0.0271 = 3.4\,\si{\joule\per\kelvin}.
\]
The original warm water cools from $298.15$ to $T_f$, losing
\[
\Delta S_{\text{cool}} = m_w c \ln\frac{T_f}{298.15}
= 0.150 \times 4180 \times \ln\frac{280.7}{298.15}
= 627 \times (-0.0604) = -37.9\,\si{\joule\per\kelvin}.
\]
The total entropy change of the isolated glass is
\[
\Delta S_{\text{tot}} = 36.7 + 3.4 - 37.9 = +2.2\,\si{\joule\per\kelvin} > 0,
\]
positive, the spontaneous direction the second law demands. The melting and
warming of the cold ice generate more entropy than the cooling of the warm water
removes, because the heat that crosses from the warm water to the ice is divided
by the ice's lower temperature into more entropy than it carried at the warm
water's higher temperature. The example also flags a numerical discipline: the
positive total is the small difference of two large terms near
$37\,\si{\joule\per\kelvin}$, so the calculation must carry the logarithms to
enough figures that the residual survives, because the second law lives in that
residual. Against a $T_0 = 298\,\si{\kelvin}$ environment the process destroys
$T_0\,\Delta S_{\text{tot}} = 298\times 2.2 \approx 0.66\,\text{kJ}$ of work
potential, the price of letting the ice melt by mixing rather than by running the
temperature difference through an engine.

\subsection{Adiabatic demagnetization: cooling by unloading spin entropy}

The lowest temperatures in a laboratory are reached not by pumping heat but by
manipulating entropy directly, and the cleanest instance is the cooling of a
paramagnetic salt by adiabatic demagnetization. We carry the two-stroke cycle
through with a spin-$\tfrac12$ model, compute the temperature drop, and connect
the result to the third law's unattainability statement.

The working substance is a mole of independent spin-$\tfrac12$ moments in a
magnetic field $B$. Each spin has two states, aligned with the field at energy
$-\mu B$ and against it at $+\mu B$, so a spin is a two-level system with gap
$\epsilon = 2\mu B$. Its spin entropy per mole is the two-level entropy of
section~4.2, a function of the single ratio $x = k_B T/(2\mu B)$: at high field
or low temperature the spins align, $x \to 0$, and the spin entropy falls toward
zero, while at low field or high temperature the spins randomise, $x \to \infty$,
and the entropy climbs toward its ceiling $R\ln 2 = 5.76\,\si{\joule\per\kelvin}$
per mole. The entropy depends on $T$ and $B$ only through their ratio, which is
the entire lever the method turns.

Stroke one magnetizes the salt at a fixed temperature. Hold the salt in thermal
contact with a bath at $T_i = 1.0\,\si{\kelvin}$ and raise the field from a small
$B_1$ to a large $B_2$. Raising the field aligns the spins and drops the spin
entropy; the shed entropy leaves as heat to the bath, which carries it away. The
salt ends stroke one cold and ordered, at the same $T_i$ but with its spins
combed into line. Stroke two removes the field adiabatically. With the salt now
thermally isolated, lowering the field from $B_2$ back to $B_1$ exchanges no heat,
so the total entropy is fixed. The spin entropy alone would rise as the field
falls, but with no heat in, the lattice must give up the difference: the spins
disorder at the expense of the lattice vibrations, and the whole salt cools.
Because the spin entropy is a function of $T/B$, holding the spin entropy fixed
holds $T/B$ fixed, so the final temperature scales with the field,
\[
T_f = T_i\,\frac{B_1}{B_2}.
\]
Starting at $T_i = 1.0\,\si{\kelvin}$ with $B_2 = 1.0\,\text{T}$ and demagnetizing
to a residual $B_1 = 10\,\text{mT}$ (the field never reaches exactly zero, because
internal dipole fields set a floor) gives $T_f = 1.0\times 0.01 = 0.01\,
\si{\kelvin} = 10\,\text{mK}$, a hundredfold cooling in one stroke.
Figure~\ref{fig:vol04ch04:adiabatic-demag} draws the two strokes on the spin
entropy-temperature plane: a vertical drop as the salt magnetizes at fixed
temperature, then a horizontal slide along a constant-entropy line back to the
low-field curve at a lower temperature. The staircase reaches toward absolute zero
but never touches it, because the low-field and high-field entropy curves converge
as $T \to 0$, so each stroke buys a smaller drop than the last. That convergence
is the third law wearing its operational face: no finite sequence of
magnetize-demagnetize strokes reaches $T = 0$, the same unattainability the
heat-capacity argument of section~4.6 delivered from a different direction.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-adiabatic-demag.tex}

\subsection{The entropy of a photon gas and the quality of radiation}

Radiation carries entropy as surely as a gas of molecules does, and quantifying
it settles what makes sunlight a high-quality energy source and waste heat a
low-quality one. We derive the entropy of blackbody radiation from its energy
density and read the quality of a radiation stream off the ratio of its energy to
its entropy.

The energy density of blackbody radiation in a cavity at temperature $T$ is the
Stefan-Boltzmann result $u = a T^4$, with $a = 7.566\times10^{-16}\,\text{J}/
(\text{m}^3\cdot\text{K}^4)$ the radiation constant. The entropy density follows
from the thermodynamic relation for a system whose energy is fixed by temperature
alone at constant volume, $ds = (1/T)\,du = (1/T)\,(4 a T^3)\,dT = 4 a T^2\,dT$,
which integrates from zero to
\[
s = \int_0^T 4 a T'^2\,dT' = \frac{4}{3} a T^3.
\]
The entropy density of radiation scales as the cube of temperature while the
energy density scales as the fourth power, so their ratio,
\[
\frac{u}{s} = \frac{a T^4}{\tfrac43 a T^3} = \frac{3}{4}\,T,
\]
grows linearly with temperature. Figure~\ref{fig:vol04ch04:photon-gas-entropy}
plots both densities on a logarithmic axis, where the two power laws read off as
straight lines of slope four and slope three. The linear rise of $u/s$ with
temperature is the working definition of radiation quality: a joule of radiation
from a hot source carries less entropy, so more of it can be turned into work,
than a joule from a cold source. Sunlight from a $T_{\text{sun}} = 5800\,
\si{\kelvin}$ photosphere carries entropy $u/s$ at $\tfrac34\times 5800 = 4350$
in $k_B$-scaled units per joule; the infrared the Earth radiates to a $255\,
\si{\kelvin}$ sky carries entropy at $\tfrac34\times 255 = 191$, twenty-three
times more entropy per joule. That factor is the thermodynamic headroom the
biosphere and the weather live inside, developed in the planetary vignette below,
and it is the reason a solar panel or a leaf is bounded not by the Carnot factor
of the ambient but by the far higher quality of the incoming photons.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-photon-gas-entropy.tex}

\subsection{Osmotic pressure as an entropic force}

The pressure that drives water across a membrane into a salt solution is entropic
in origin, the same mixing entropy that drove the free expansion, and computing it
shows a bulk mechanical force emerging from a count of configurations. We derive
the van't Hoff osmotic law from the chemical potential and evaluate it for
seawater and for a living cell.

A membrane permeable to solvent but not to solute separates pure solvent from a
dilute solution. The solvent's chemical potential in the solution is lowered by
the presence of the solute, because adding solute raises the mixing entropy and so
lowers the solvent's free energy per molecule. For a dilute solution of mole
fraction $x_s$ of solute, the solvent chemical potential falls by $\mu_{\text{solv}}
= \mu^\circ + RT\ln(1 - x_s) \approx \mu^\circ - RT x_s$. Solvent flows across the
membrane toward the lower chemical potential, into the solution, and it keeps
flowing until a hydrostatic pressure difference, the osmotic pressure $\Pi$, raises
the solution-side chemical potential by $V_m\Pi$ back to balance. Setting the
raised potential equal on the two sides, $V_m\Pi = RT x_s$, and using $x_s \approx
n_s V_m/V$ for a dilute solution gives the van't Hoff law
\[
\Pi = \frac{n_s}{V}\,RT = c_s R T,
\]
with $c_s$ the solute molar concentration. The form is the ideal-gas law with the
solute playing the role of a gas that fills the solution volume, because the
driving entropy is the same mixing entropy in both cases. For seawater, with a
total dissolved-ion concentration of about $c_s = 1.1\,\text{mol}/\text{L} =
1100\,\text{mol}/\text{m}^3$ at $T = 290\,\si{\kelvin}$,
\[
\Pi = 1100 \times 8.314 \times 290 = 2.65\times10^{6}\,\text{Pa} \approx 26\,
\text{bar},
\]
which is the pressure a reverse-osmosis desalination plant must exceed to push
fresh water back out of the brine, and the exergy of that separation is $\Pi$
times the permeate volume, the thermodynamic floor on the energy cost of
desalination. For a red blood cell, whose interior is roughly $0.3\,\text{mol}/
\text{L}$ more concentrated than pure water, the osmotic pressure across the
membrane is $\Pi = 300\times 8.314\times 310 \approx 7.7\times10^{5}\,\text{Pa}
\approx 7.6\,\text{bar}$, the pressure the cell would burst under in fresh water
and the reason intravenous fluids are matched to blood's tonicity. A count of
mixing configurations, carried through the chemical potential, sets a pressure an
engineer must design a membrane, a pump, or a cell wall to hold.

\subsection{Sackur-Tetrode entropy evaluated for argon}

The Sackur-Tetrode formula is derived in the exercises; here we put numbers
through it and check the microscopic prediction against a tabulated standard
entropy, one of the sharpest quantitative tests that the microstate count is
right. We compute the absolute molar entropy of argon gas at standard conditions
from the formula and compare to the calorimetric value.

The Sackur-Tetrode entropy of a monatomic ideal gas is
\[
S = N k_B\left[\ln\!\left(\frac{V}{N\Lambda^3}\right) + \frac{5}{2}\right],
\qquad
\Lambda = \frac{h}{\sqrt{2\pi m k_B T}},
\]
with $\Lambda$ the thermal de Broglie wavelength. For argon at $T = 298.15\,
\si{\kelvin}$ and $p = 1\,\text{bar}$, the atomic mass is $m = 39.95\times
1.6605\times10^{-27} = 6.634\times10^{-26}\,\text{kg}$. The thermal wavelength is
\[
\Lambda = \frac{6.626\times10^{-34}}
{\sqrt{2\pi\times 6.634\times10^{-26}\times 1.381\times10^{-23}\times 298.15}}
= 1.60\times10^{-11}\,\text{m},
\]
about a sixtieth of an atomic diameter, small enough that the classical count is
valid. The volume per particle at one bar is $V/N = k_B T/p = 1.381\times10^{-23}
\times 298.15/10^{5} = 4.12\times10^{-26}\,\text{m}^3$, so the quantum
concentration ratio is
\[
\frac{V}{N\Lambda^3} = \frac{4.12\times10^{-26}}{(1.60\times10^{-11})^3}
= \frac{4.12\times10^{-26}}{4.10\times10^{-33}} = 1.00\times10^{7}.
\]
The molar entropy is then
\[
S = R\left[\ln(1.00\times10^{7}) + \frac52\right]
= 8.314\,(16.12 + 2.50)
= 8.314\times 18.62 = 154.8\,\si{\joule\per\kelvin}\ \text{per mole}.
\]
The calorimetric standard molar entropy of argon, built by integrating measured
heat capacity from near absolute zero as in the third-law example above, is
$154.8\,\si{\joule\per\kelvin}$ per mole. The two agree to the figures carried,
which is the quantitative vindication of $S = k_B\ln W$: a formula assembled from
Planck's constant, the atomic mass, and a count of quantum phase-space cells
reproduces a number a calorimeter measures without ever mentioning atoms. The
logarithm's large argument, $10^7$, is the number of thermally accessible
translational cells per atom, and the additive $\tfrac52$ splits as $\tfrac32$
from the kinetic temperature dependence and $1$ from the Gibbs
indistinguishability correction, the same $N!$ that resolved the mixing paradox
of section~4.1.

\subsection{Entropy generation in an adiabatic mixing chamber}

Two streams at different temperatures blend in an open chamber, a heater feedwater
tee or a hot-and-cold tap, and the blending generates entropy even though no heat
crosses to the outside. We compute the entropy generated when two water streams
mix and the exergy that destroys, a steady-flow companion to the closed-system ice
example.

A stream of $\dot m_1 = 2.0\,\text{kg}/\text{s}$ of water at $T_1 = 350\,
\si{\kelvin}$ mixes adiabatically at constant pressure with $\dot m_2 = 3.0\,
\text{kg}/\text{s}$ at $T_2 = 290\,\si{\kelvin}$. An energy balance sets the outlet
temperature as the mass-weighted mean, since the specific heat $c = 4.18\,
\text{kJ}/(\text{kg}\cdot\si{\kelvin})$ is common to both,
\[
T_3 = \frac{\dot m_1 T_1 + \dot m_2 T_2}{\dot m_1 + \dot m_2}
= \frac{2.0\times 350 + 3.0\times 290}{5.0}
= \frac{700 + 870}{5.0} = 314\,\si{\kelvin}.
\]
Each stream's entropy changes as it moves to the common outlet temperature. The
hot stream cools and loses entropy; the cold stream warms and gains entropy; the
net is the generation,
\[
\dot S_{\text{gen}} = \dot m_1 c \ln\frac{T_3}{T_1} + \dot m_2 c \ln\frac{T_3}{T_2}
= 2.0\times 4180 \ln\frac{314}{350} + 3.0\times 4180 \ln\frac{314}{290}.
\]
The first term is $8360\times(-0.1085) = -907\,\si{\watt\per\kelvin}$ and the
second is $12{,}540\times(0.0795) = +997\,\si{\watt\per\kelvin}$, so
\[
\dot S_{\text{gen}} = -907 + 997 = +90\,\si{\watt\per\kelvin} > 0,
\]
positive, as an irreversible mix must be. Against a $T_0 = 290\,\si{\kelvin}$
environment the exergy destroyed is $T_0\,\dot S_{\text{gen}} = 290\times 90 =
26\,\text{kW}$, the work an ideal engine could have drawn from the two streams'
temperature difference before they were allowed to blend. The mixing chamber is
the steady-flow twin of the ice glass and the heat-across-a-gap examples: a
temperature difference discarded by direct contact rather than run through an
engine, and the entropy generated is again the small positive residual of two
larger opposite-signed terms, so the logarithms must be carried to enough figures
that the residual survives. The lesson repeats across the chapter's examples, from
the counterflow exchanger to the melting ice: any time two bodies at different
temperatures are brought together without an engine between them, the work
potential in their difference is destroyed, and the amount destroyed is
$T_0$ times the entropy the blending generates.

\subsection{The heat pump and the exergy of warmth}

A resistance heater turns a joule of electrical work into a joule of
room-temperature heat, and a first-law accounting calls it perfectly
efficient. A heat pump doing the same warming job draws far less than a joule
of work per joule delivered, which sounds like a first-law violation until the
second law is brought in. We size an ideal heat pump warming a house and show
where the apparent free heat comes from.

The house is held at $T_H = 293\,\si{\kelvin}$ ($20\,\degC$) against an outdoor
source at $T_C = 273\,\si{\kelvin}$ ($0\,\degC$), and it loses heat to the cold
outside at a steady $\dot Q_H = 8\,\text{kW}$ that the heating plant must
replace. A resistance heater supplies this by dissipating $8\,\text{kW}$ of
electrical work, coefficient of performance one. An ideal heat pump instead
lifts heat from the cold outdoors and delivers it indoors, with the reversible
heating coefficient of performance
\[
\mathrm{COP}_H = \frac{T_H}{T_H - T_C} = \frac{293}{293 - 273} = \frac{293}{20}
= 14.7.
\]
The reversible work to deliver $8\,\text{kW}$ of warmth is therefore
$\dot W_{\text{rev}} = \dot Q_H/\mathrm{COP}_H = 8000/14.7 = 545\,\text{W}$, a
fifteenth of the resistance heater's draw. The other $7.45\,\text{kW}$ is heat
pumped from the outdoor air, delivered indoors for free, because the pump moves
ambient heat uphill rather than manufacturing it from work.
Figure~\ref{fig:vol04ch04:heatpump-cop} draws the coefficient against source
temperature: it climbs without bound as the source warms toward the delivery
temperature and collapses toward one as the source cools and the lift widens.
The heat pump does not beat the first law; the delivered heat is not created,
it is relocated, and the second law caps the multiplier at the Carnot value.

The exergy view sharpens the point. The work is high-quality energy, all of it
convertible; the delivered $8\,\text{kW}$ of warmth at $293\,\si{\kelvin}$
against a $T_0 = 273\,\si{\kelvin}$ dead state carries exergy only $\dot Q_H(1 -
T_0/T_H) = 8000\,(1 - 273/293) = 546\,\text{W}$, almost exactly the reversible
work in. The heat pump is an exergy-conserving device run in reverse: it spends
$545\,\text{W}$ of pure exergy to deliver $546\,\text{W}$ of exergy as warmth,
and the seven kilowatts of low-grade heat it also delivers carry almost no
exergy at all. The resistance heater, by contrast, destroys nearly all the
exergy of its electrical input, turning $8\,\text{kW}$ of pure work into
$8\,\text{kW}$ of near-dead-state heat worth only $546\,\text{W}$ of work
potential. The two devices deliver identical energy and wildly different exergy
bills, the same lesson the steam-plant audit drew, now written for a heating
appliance.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-heatpump-cop.tex}

\subsection{The Nernst equation and the concentration cell}

The fuel-cell case study read a voltage off a reaction's free energy. A
concentration cell isolates the entropy term alone: the same species sits on
both electrodes, so the enthalpy of reaction is zero, and the cell drives a
current purely on the entropy of mixing between two concentrations. We compute
the voltage of a copper concentration cell and show it is an electrochemical
free expansion.

Two copper electrodes dip into copper-sulfate solutions of different
concentrations, $c_1 = 0.010\,\text{mol}/\text{L}$ on one side and $c_2 =
1.0\,\text{mol}/\text{L}$ on the other, joined by a salt bridge. Copper
dissolves from the dilute side and plates onto the concentrated side, the
direction that carries ions from high concentration toward low and so raises
the mixing entropy. The reaction is $\text{Cu}^{2+}(c_2) \to \text{Cu}^{2+}(c_1)$
in net effect, with standard free energy zero because the species is identical
on both sides; the whole driving force is the concentration ratio. The Nernst
equation gives the cell voltage,
\[
E = -\frac{RT}{nF}\ln\frac{c_1}{c_2}
= \frac{RT}{nF}\ln\frac{c_2}{c_1},
\]
with $n = 2$ electrons per copper ion and $F = 96{,}485\,\text{C}/\text{mol}$.
At $T = 298\,\si{\kelvin}$,
\[
E = \frac{8.314 \times 298}{2 \times 96{,}485}\ln\frac{1.0}{0.010}
= 0.01284 \times \ln 100
= 0.01284 \times 4.605
= 0.059\,\text{V}.
\]
The cell holds fifty-nine millivolts, the textbook $59\,\text{mV}$ per decade of
concentration ratio at room temperature that underlies the pH electrode and the
ion-selective sensor. As the current flows the two concentrations converge, the
ratio falls toward one, and the voltage decays to zero when the concentrations
equalise. That endpoint is the electrochemical version of the free expansion of
section~4.1: the system runs until the mixing entropy is maximised, and the
work it delivers on the way, $nFE$ per mole, is exactly $RT\ln(c_2/c_1)$, the
same $RT\ln(V_2/V_1)$ that a gas delivers expanding into a larger volume. A
concentration difference is stored exergy, and the concentration cell is the
device that spends it as electrical work rather than discarding it to mixing.

\subsection{Effusion, Graham's law, and isotope separation}

The Maxwell-Boltzmann distribution is not only a picture; it sets the rate at
which a gas escapes through a small hole and the degree to which that escape
sorts a mixture by mass. We compute the effusion rate and show why the same
mass dependence that makes light molecules escape faster is the thermodynamic
basis of uranium enrichment.

A gas at number density $n$ and temperature $T$ strikes a wall at the rate
$\tfrac14 n \bar v$ per unit area, with $\bar v = \sqrt{8 k_B T/\pi m}$ the mean
molecular speed from the Maxwell-Boltzmann distribution. A small hole of area
$A$, small enough that the gas near it stays in equilibrium, lets molecules
effuse at
\[
\frac{dN}{dt} = \frac14 n \bar v A = \frac{n A}{4}\sqrt{\frac{8 k_B T}{\pi m}}
= n A \sqrt{\frac{k_B T}{2\pi m}}.
\]
The rate scales as $m^{-1/2}$: a lighter molecule effuses faster, in the ratio
$\sqrt{m_2/m_1}$ for two species at the same temperature and density, which is
Graham's law. For a mixture the escaping stream is enriched in the lighter
species by the factor $\sqrt{m_{\text{heavy}}/m_{\text{light}}}$ per stage. Take
uranium hexafluoride, the gas of the enrichment industry, with $^{235}\text{UF}_6$
of molar mass $349$ and $^{238}\text{UF}_6$ of molar mass $352$. The
single-stage separation factor is
\[
\alpha = \sqrt{\frac{352}{349}} = \sqrt{1.00859} = 1.00429,
\]
a gain of four parts in a thousand per stage. Enriching natural uranium from its
$0.72\%$ of $^{235}\text{U}$ to the few percent a reactor needs, or the high
fraction a weapon needs, therefore takes a cascade of many hundreds to thousands
of effusion stages in series, each nudging the ratio by $0.4\%$, which is why
gaseous-diffusion enrichment plants were among the largest and most
energy-hungry industrial facilities ever built. The thermodynamic root of that
scale is the square-root mass dependence of the Maxwell-Boltzmann speed: the
mass difference between the isotopes is tiny, so the per-stage sorting the
distribution provides is tiny, and the entropy the cascade fights, the mixing
entropy that wants the isotopes blended, must be paid down one small step at a
time.

\subsection{Trouton's rule and the entropy of vaporization}

The latent heat of a liquid and its boiling point are not independent; the
entropy of vaporization ties them together, and its near-constancy across
liquids is a statement about the disorder a molecule gains on leaving the
liquid. We state Trouton's rule, test it, and read its exceptions.

At the normal boiling point a liquid and its vapour coexist in equilibrium, so
the free energy of vaporization is zero and the entropy of vaporization is the
latent heat divided by the boiling temperature,
\[
\Delta S_{\text{vap}} = \frac{\Delta H_{\text{vap}}}{T_b}.
\]
Trouton's rule is the observation that $\Delta S_{\text{vap}} \approx
88\,\si{\joule\per\kelvin}$ per mole for a wide range of liquids, because the
disorder a molecule gains in going from the close-packed liquid to the dilute
vapour is roughly the same regardless of the substance. For benzene, with
$\Delta H_{\text{vap}} = 30.8\,\text{kJ}/\text{mol}$ and $T_b = 353\,\si{\kelvin}$,
\[
\Delta S_{\text{vap}} = \frac{30{,}800}{353} = 87.3\,\si{\joule\per\kelvin}
\ \text{per mole},
\]
squarely on the rule. For argon, $\Delta H_{\text{vap}} = 6.5\,\text{kJ}/\text{mol}$
and $T_b = 87\,\si{\kelvin}$, giving $6500/87 = 74.7\,\si{\joule\per\kelvin}$,
close. Water breaks the rule: $\Delta H_{\text{vap}} = 40.7\,\text{kJ}/\text{mol}$
and $T_b = 373\,\si{\kelvin}$ give $40{,}700/373 = 109\,\si{\joule\per\kelvin}$,
well above $88$, because hydrogen bonding orders the liquid more than a normal
liquid, so a water molecule gains extra entropy on escaping into the vapour.
The rule and its exceptions are read directly off the Clausius-Clapeyron plot of
figure~\ref{fig:vol04ch04:clausius-clapeyron}: the slope of $\ln p$ against
$1/T$ is $-\Delta H_{\text{vap}}/R$, and the same latent heat divided by the
boiling temperature is the entropy of vaporization Trouton's rule bounds. A
liquid that hydrogen-bonds, like water or an alcohol, sits above the Trouton
line; a liquid whose molecules cluster in the vapour, like a carboxylic acid,
falls below it. The entropy of vaporization is a fingerprint of how ordered the
liquid was before it boiled.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-clausius-clapeyron.tex}

\subsection{The isothermal atmosphere and the barometric distribution}

The Boltzmann factor governs occupation of energy levels, and gravitational
potential energy is a level like any other. Placing an ideal gas in a
gravitational field and asking for its equilibrium density profile turns the
Boltzmann distribution into the barometric law and shows a bulk pressure
gradient emerging from a microscopic weighting. We derive the profile and
evaluate a scale height.

A column of ideal gas at uniform temperature $T$ in gravity $g$ has each
molecule carrying potential energy $mgz$ at height $z$. The equilibrium
probability of finding a molecule at height $z$ is the Boltzmann factor with
this energy, so the number density follows
\[
n(z) = n(0)\,e^{-mgz/k_B T},
\]
the barometric distribution. The same result follows from mechanical
equilibrium: a slab of gas of thickness $dz$ is held up by the pressure
difference across it, $dp = -\rho g\,dz$, and substituting the ideal-gas
$\rho = mp/k_B T$ integrates to the same exponential. The two derivations agree
because the pressure gradient a fluid engineer computes from force balance and
the density profile a statistical mechanic computes from the Boltzmann factor
are the same physics read two ways. The falloff length is the scale height
\[
H = \frac{k_B T}{mg} = \frac{RT}{Mg},
\]
the altitude over which the density drops by a factor $e$. For nitrogen at $T =
250\,\si{\kelvin}$, $H = 8.314 \times 250/(0.028 \times 9.81) = 7.6\,\text{km}$,
so the air thins to a third of its sea-level density near the cruising altitude
of an airliner. Figure~\ref{fig:vol04ch04:barometric-distribution} plots the
profile for three species: the heavier the molecule, the shorter the scale
height and the faster the density thins with altitude, because gravity pulls
harder on a heavier molecule while the thermal energy that spreads it stays the
same. That mass dependence is why a centrifuge, which replaces $g$ with a
spin-induced acceleration thousands of times larger, sorts isotopes far more
effectively per stage than gaseous effusion, and why the lightest gases, helium
and hydrogen, leak steadily off the top of the atmosphere while nitrogen and
oxygen stay bound. The equilibrium profile is the balance the second law
strikes between the entropy the gas gains in spreading upward and the energy it
costs to climb.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-barometric-distribution.tex}

\subsection{Bose-Einstein and Fermi-Dirac occupation, and when the classical count holds}

Section~4.2 built the canonical thermodynamics on a classical count of states
and flagged that the count fails when quantum effects matter. We make the
boundary quantitative: identical particles obey one of two quantum statistics,
and the classical Boltzmann count is the common dilute limit of both. Knowing
where that limit breaks tells an engineer when the handbook tables need a
quantum correction.

Identical quantum particles come in two kinds. Bosons, with integer spin, may
share a single-particle state without limit; fermions, with half-integer spin,
forbid double occupancy by the Pauli principle. Counting the microstates of
each under the constraint of fixed average particle number gives the mean
occupation of a single-particle state of energy $\epsilon$,
\[
\langle n\rangle_{\text{BE}} = \frac{1}{e^{(\epsilon-\mu)/k_B T} - 1},
\qquad
\langle n\rangle_{\text{FD}} = \frac{1}{e^{(\epsilon-\mu)/k_B T} + 1},
\]
where $\mu$ is the chemical potential fixed by the total particle number. The
Fermi-Dirac occupation never exceeds one, the exclusion principle built into
the sign in the denominator, and at $\epsilon = \mu$ it equals one half. The
Bose-Einstein occupation grows without bound as $\epsilon$ approaches $\mu$ from
above, the crowding that produces Bose-Einstein condensation.
Figure~\ref{fig:vol04ch04:quantum-statistics} draws both against the classical
Maxwell-Boltzmann form $\langle n\rangle_{\text{MB}} = e^{-(\epsilon-\mu)/k_B T}$.
When the occupation is small, $\langle n\rangle \ll 1$, the exponential in the
denominator dominates the $\pm 1$, and both quantum curves collapse onto the
classical exponential. The dilute condition is that the thermal de Broglie
wavelength be much smaller than the interparticle spacing,
\[
\frac{n\Lambda^3}{1} \ll 1,
\qquad \Lambda = \frac{h}{\sqrt{2\pi m k_B T}}.
\]
For air at room temperature $n\Lambda^3 \sim 10^{-7}$, so the classical count
of section~4.2 is excellent and the Sackur-Tetrode entropy holds. For the
conduction electrons of a metal, light and dense, $n\Lambda^3 \sim 10^{4}$ at
room temperature, so the electron gas is deeply degenerate and its heat capacity
is the linear-in-$T$ Fermi term of section~4.6, not the classical
$\tfrac32 k_B$ per electron. For liquid helium below a few kelvin the boson
count produces superfluidity. The one dimensionless group $n\Lambda^3$ decides
which world an engineer is in: dilute and classical, or dense and quantum.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-quantum-statistics.tex}

\subsection{Mutual information and the capacity of a noisy channel}

The information-entropy preview named channel capacity as a mutual information
without computing one. We carry a binary symmetric channel through in full,
because the same relative-entropy machinery that underwrites the Boltzmann
distribution sets the ceiling on how fast a sensor or a communication link can
report.

A binary symmetric channel carries a bit that arrives flipped with probability
$p$ and intact with probability $1 - p$. The mutual information between the sent
symbol $X$ and the received symbol $Y$ is the entropy of the output less the
entropy left after the input is known,
\[
I(X;Y) = H(Y) - H(Y \mid X).
\]
The noise sets the conditional term: given the input, the output is wrong with
probability $p$, so $H(Y\mid X) = H_2(p)$, the binary entropy function $H_2(p) =
-p\log_2 p - (1-p)\log_2(1-p)$. The output entropy $H(Y)$ is maximised at one
bit when the input is equiprobable, so the channel capacity is
\[
C = 1 - H_2(p),
\]
in bits per use. For a clean channel, $p = 0$, the capacity is one bit per use,
the whole symbol gets through. For a channel that flips half the bits, $p = 0.5$,
the binary entropy is one and the capacity is zero: a symbol arriving right half
the time and wrong half the time carries no information, because the output is
independent of the input. For a realistic $p = 0.01$,
\[
C = 1 - [-0.01\log_2 0.01 - 0.99\log_2 0.99]
= 1 - [0.0664 + 0.0144]
= 1 - 0.0808 = 0.919\ \text{bits per use},
\]
so a one-percent error rate costs about eight percent of the channel's
capacity, and Shannon's theorem promises that codes exist which recover the
transmitted message with vanishing error at any rate below $0.919$ bits per use.
The capacity is a mutual information, the mutual information is a difference of
Shannon entropies, and the Shannon entropy is the same quantity, up to the
factor $k_B$, that section~4.1 counted as thermodynamic entropy. An engineer
sizing a sensor's bit rate against its noise, or a link's throughput against its
error rate, is reading a ceiling set by the same entropy that governs a gas.

\subsection{The Gough-Joule effect: a rubber band that heats on stretch}

The polymer worked example derived the entropic spring; its thermal signature is
a rubber band that warms when stretched and cools when released, the reverse of
a metal wire. We compute the temperature change and connect it to the sign of
the entropy's dependence on extension.

A rubber band under tension $f$ at extension $L$ obeys the thermodynamic
identity, from the first and second laws applied to a one-dimensional elastic,
\[
\left(\pdv{T}{L}\right)_S = -\frac{T}{C_L}\left(\pdv{f}{T}\right)_L,
\]
an elastic analogue of the Maxwell relations, where $C_L$ is the heat capacity
at fixed length. For the entropic spring the tension is $f = (3 k_B T/N b^2)\,R$,
proportional to temperature, so $(\partial f/\partial T)_L > 0$. The identity
then makes $(\partial T/\partial L)_S < 0$ read the other way: stretching at
constant entropy, done fast so no heat escapes, raises the temperature, because
the tension rises with temperature. Equivalently, the entropy falls with
extension, $(\partial S/\partial L)_T < 0$, so an adiabatic stretch that holds
total entropy fixed must raise the temperature to compensate the lost
configurational entropy. Numerically, a rubber band stretched quickly to double
its length can warm by a degree or two, easily felt by pressing it to the lip.
Releasing it reverses the sign: the band cools as it snaps back, the
configurational entropy climbing again at the expense of the thermal energy.
This is the Gough-Joule effect, and it is the exact opposite of a metal wire,
which cools on adiabatic stretch because its elasticity is energetic, not
entropic. The direction of the temperature change is a direct readout of the
sign of $(\partial S/\partial L)_T$, and it is why rubber, alone among common
solids, has a negative coefficient of thermal expansion under load: a stretched
band contracts when heated, pulling harder as its entropy fights to recoil the
chains into their many coiled states.

\subsection{Residual entropy: why ice keeps entropy at absolute zero}

The third law fixes the entropy of a system with a unique ground state at zero,
and the third-law worked example noted that a degenerate ground state leaves a
residual. Ice is the classic case, and Pauling's count of it is one of the
cleanest applications of $S = k_B \ln W$ in the whole subject. We reproduce the
count and compare it to the measured residual entropy of ice.

Water ice is a lattice of oxygen atoms with a hydrogen between each adjacent
pair, but the hydrogen is not centred: it sits closer to one oxygen than the
other, and each oxygen must have exactly two near hydrogens (the two covalent
bonds of the water molecule) and two far ones (the two hydrogen bonds it
accepts). These are the ice rules. Even at absolute zero the hydrogens can
arrange themselves in enormously many ways consistent with the rules, so the
ground state is degenerate and the entropy does not vanish. Pauling's count runs
as follows. Each of the $N$ oxygen atoms has four bonds, and each bond has two
placements for its hydrogen, giving $2^{2N} = 4^N$ raw configurations. Of the
sixteen hydrogen arrangements around a single oxygen, only six satisfy the
two-near-two-far rule, a fraction $6/16 = 3/8$. Treating the constraints at
different oxygens as independent, the surviving count is
\[
W = 4^N \left(\frac{3}{8}\right)^N = \left(\frac{3}{2}\right)^N,
\]
so the residual entropy per mole is
\[
S_0 = k_B \ln W = N_A k_B \ln\frac{3}{2} = R\ln 1.5
= 8.314 \times 0.405 = 3.37\,\si{\joule\per\kelvin}\ \text{per mole}.
\]
The measured residual entropy of ice, the gap between its calorimetric entropy
integrated from near absolute zero and the Sackur-Tetrode entropy of water
vapour computed spectroscopically, is $3.4\,\si{\joule\per\kelvin}$ per mole,
matching Pauling's count to the figures carried. The agreement is a second
quantitative triumph of $S = k_B \ln W$, alongside the argon calculation: a
combinatorial count of proton arrangements, carried out on the back of an
envelope, reproduces a number two independent measurements pin down. The
residual entropy is the count's way of saying that ice does not reach a unique
ground state, so the third law's zero applies to the ordered crystal it never
quite forms, and the frozen-in disorder of the protons is entropy the crystal
carries down to absolute zero.

\subsection{The thermodynamic temperature scale from a Carnot cascade}

Temperature has been used throughout as a number a thermometer reads, but the
second law defines it without any thermometer fluid at all, and pinning down that
definition removes the last arbitrary constant from the entropy. We show that a
reversible engine's heat ratio depends on the reservoir temperatures alone, use
that to build the Kelvin scale, and locate absolute zero as the point where the
engine rejects no heat.

A reversible engine exchanging heat $Q_H$ with a hot reservoir and $Q_C$ with a
cold one has a heat ratio $Q_H/Q_C$ that cannot depend on the working substance,
because two reversible engines between the same two reservoirs with different heat
ratios could be coupled, one driving the other backward, to move heat from cold to
hot with no work, violating the Clausius statement. The ratio therefore depends on
the two reservoir temperatures only, $Q_H/Q_C = f(T_H, T_C)$ for some universal
function. Running a reversible engine from $T_1$ to $T_2$ and a second from $T_2$
to $T_3$ in cascade, so the second absorbs exactly what the first rejects, forces
$f(T_1, T_3) = f(T_1, T_2)\,f(T_2, T_3)$, and the only functions with that
multiplicative property factor as $f(T_H, T_C) = \phi(T_H)/\phi(T_C)$. Kelvin's
choice is to define the thermodynamic temperature as $\phi$ itself, so that
\[
\frac{Q_H}{Q_C} = \frac{T_H}{T_C},
\]
the defining relation drawn as the straight line of
figure~\ref{fig:vol04ch04:kelvin-scale}. The scale is fixed up to one multiplicative
constant, set historically by assigning the triple point of water $273.16$ and now
by fixing the Boltzmann constant, and it is the scale in which the Carnot efficiency
takes the form $1 - T_C/T_H$ used everywhere in this volume. To calibrate a second
fixed point, run a reversible engine between the reference reservoir and the unknown
and measure the heat ratio: an engine drawing $Q_{\text{ref}} = 100\,\text{J}$ from
a reservoir at the reference $273.16\,\si{\kelvin}$ and rejecting $Q = 73.2\,\text{J}$
to an unknown cold reservoir places that reservoir at $T = 273.16 \times 73.2/100 =
200\,\si{\kelvin}$, no thermometer fluid consulted. Absolute zero is the temperature
at which a reversible engine converts all its intake to work and rejects $Q_C = 0$,
the origin where the line meets the axis; the third law's unattainability is the
statement that this rejection can be driven toward zero but never reached. The
thermodynamic scale is what makes $S = \int \delta Q_{\text{rev}}/T$ a definite
quantity rather than a fluid-dependent one, because the $T$ in the denominator is
now defined by the same reversible engine that defines entropy.

\subsection{The detailed-balance limit of a solar cell}

The photon-gas example measured the quality of sunlight; a solar cell is the device
that cashes that quality into electrical work, and the second law sets a ceiling on
how much it can cash. We compute the two thermodynamic ceilings on solar conversion
and then the detailed-balance limit of a single-threshold cell, to show why a real
photovoltaic peaks near a third and not near the radiation ceiling.

Sunlight arriving from a $T_{\text{sun}} = 5800\,\si{\kelvin}$ source and converted
against a $T_0 = 300\,\si{\kelvin}$ sink is bounded, if treated as heat from a hot
reservoir, by the Carnot factor $1 - T_0/T_{\text{sun}} = 1 - 300/5800 = 0.948$.
A slightly tighter bound accounts for the entropy that must be generated in
reshaping the broadband solar spectrum into work, the Landsberg limit, which sits
near $0.933$ for these temperatures; the small gap between the two is the entropy
tax on spectral conversion. A real cell cannot reach either, because it absorbs
through a single threshold. A semiconductor of band gap $E_g$ absorbs every photon
with energy above $E_g$ and none below it, and each absorbed photon delivers at most
$E_g$ of usable energy regardless of how far above the gap it arrived, the excess
thermalizing to the band edge as heat. The two losses oppose each other in $E_g$: a
low gap captures many photons but wastes most of each one's energy, a high gap uses
each photon efficiently but captures few. The detailed-balance calculation, which
balances the photon absorption rate against the radiative recombination the cell
must emit as a body at $T_0$, gives an efficiency that peaks near $E_g \approx
1.34\,\si{\electronvolt}$ at about $33\%$, drawn as the interior curve of
figure~\ref{fig:vol04ch04:solar-cell-limit}. Silicon, with $E_g = 1.12\,
\si{\electronvolt}$, sits close to the peak, which is the thermodynamic reason it
dominates the industry. The gap between the $33\%$ peak and the $95\%$ Carnot
ceiling is the entropy a one-threshold device generates by thermalizing each photon
to the band edge, and it is the ceiling that multi-junction cells chase by stacking
several gaps to slice the spectrum, each junction converting a narrower band closer
to its own edge. The lesson repeats the photon-gas example in reverse: the quality
of the incoming radiation sets the ceiling, and every irreversible step between
absorption and the terminals, thermalization first among them, is entropy generated
and work forgone.

\subsection{Entropy generation in a resistor dissipating electrical work}

The Johnson-noise discussion showed a resistor fluctuating in equilibrium; a
resistor carrying a current is doing something else, turning ordered electrical work
wholly into heat, and it is the purest entropy generator in the chapter. We compute
the entropy a loaded resistor generates and the exergy it destroys, and note that
its first-law efficiency and its second-law efficiency could not differ more.

A resistor $R = 100\,\Omega$ carries a steady current $I = 2.0\,\text{A}$, so it
dissipates electrical work at the rate $\dot W = I^2 R = 4.0 \times 100 =
400\,\text{W}$. In steady state the resistor's temperature is constant, so all
$400\,\text{W}$ of work leaves as heat to the surroundings. Take the resistor held
at a surface temperature $T = 350\,\si{\kelvin}$ by that heat flow, shedding to an
environment at $T_0 = 300\,\si{\kelvin}$. The work carries no entropy, being ordered
energy, while the heat leaving the resistor at $T$ carries entropy at the rate
$\dot Q/T = 400/350 = 1.14\,\si{\watt\per\kelvin}$, so the resistor generates
entropy at
\[
\dot S_{\text{gen}} = \frac{\dot W}{T} = \frac{400}{350}
= 1.14\,\si{\watt\per\kelvin}
\]
in converting pure work to heat at its own temperature, and the surroundings then
generate a further $400(1/300 - 1/350) = 0.19\,\si{\watt\per\kelvin}$ as the heat
slides down the remaining gap to the environment. The total entropy generation from
electrical supply to environment is $400/300 = 1.33\,\si{\watt\per\kelvin}$, and the
exergy destroyed is $T_0\,\dot S_{\text{gen}} = 300 \times 1.33 = 400\,\text{W}$,
the entire electrical input. A resistor is a device whose first-law efficiency is
one, every joule of work accounted for as heat, and whose second-law efficiency is
zero, every joule of work potential destroyed, because the input was pure exergy and
the output is near-dead-state heat worth almost nothing. The calculation is why
resistance heating, though perfectly efficient by the energy ledger, is the most
wasteful way to warm a space by the exergy ledger, the point the heat-pump example
made from the other side: the heat pump spends a fifteenth of the work to deliver
the same warmth because it moves ambient heat rather than manufacturing it from pure
work as the resistor does.

\subsection{Freezing-point depression as an entropic effect}

Salt lowers the freezing point of water, and the mechanism is the mixing entropy of
section~4.1 acting on the solvent's chemical potential, the same entropy that drove
the free expansion now moving a phase boundary. We derive the freezing-point
depression from the equality of chemical potentials and evaluate it for the salt on
a winter road.

At the freezing point the solid solvent and the liquid solution are in equilibrium,
so the solvent's chemical potential is equal in the two phases,
$\mu_{\text{solid}}(T) = \mu_{\text{liquid}}(T)$. Dissolving a solute in the liquid
lowers the solvent's chemical potential there by the mixing term $RT\ln x_{\text{solv}}
\approx -RT x_s$ for a dilute solution of solute mole fraction $x_s$, because adding
solute raises the liquid's entropy and so lowers the solvent's free energy per
molecule. The pure solid is untouched, so the balance that held at the pure freezing
point $T_f^0$ is broken, and the system must cool to a lower temperature to restore
it, where the temperature dependence of the two potentials, carrying the entropy of
fusion $\Delta S_{\text{fus}} = \Delta H_{\text{fus}}/T_f^0$, closes the gap.
Setting the shifted potentials equal to first order gives the freezing-point
depression
\[
\Delta T_f = \frac{R (T_f^0)^2}{\Delta H_{\text{fus}}}\,x_s
= K_f\, m,
\]
with $m$ the molality and $K_f = R(T_f^0)^2 M_{\text{solv}}/\Delta H_{\text{fus}}$
the cryoscopic constant, $1.86\,\si{\kelvin}\,\text{kg}/\text{mol}$ for water. Take
a road salted to a $1.0\,\text{molal}$ solution of sodium chloride, which
dissociates into two ions per formula unit, so the effective solute molality is
$2.0\,\text{mol}/\text{kg}$. The depression is
\[
\Delta T_f = 1.86 \times 2.0 = 3.7\,\si{\kelvin},
\]
so the salted water stays liquid down to about $-3.7\,\degC$, and heavier salting to
the eutectic pushes the depression to roughly $-21\,\degC$ for sodium chloride
before the salt itself begins to precipitate. The depression is a colligative
property: it depends on the number of dissolved particles and not their identity,
because the driving quantity is the mixing entropy, which counts particles rather
than weighing them. The same entropic mechanism raises the boiling point, lowers the
vapour pressure, and drives the osmotic pressure of the earlier example, four faces
of one fact, that dissolving a solute raises the solvent's entropy and so lowers its
escaping tendency from the liquid. A count of mixing configurations, carried through
the chemical potential, sets the temperature at which a winter road stays wet.

\subsection{The two-level system with degenerate levels}

The two-level system of section~4.2 took both levels as single states; real level
schemes carry degeneracies, and folding them in shifts the Schottky peak and the
high-temperature entropy in a way an engineer sizing a magnetic refrigerant or
reading a defect population must account for. We redo the two-level thermodynamics
with a lower level of degeneracy $g_0$ and an upper level of degeneracy $g_1$, and
show how the ratio moves the results.

The partition function counts each level weighted by its degeneracy,
\[
Z = g_0 + g_1 e^{-\epsilon/k_B T},
\]
so the upper-state occupation and the average energy carry the degeneracy ratio
$r = g_1/g_0$,
\[
p_1 = \frac{r e^{-\epsilon/k_B T}}{1 + r e^{-\epsilon/k_B T}},
\qquad
\langle E\rangle = \epsilon\,p_1.
\]
At high temperature the exponential goes to one and the occupation saturates not at
one half but at $p_1 \to r/(1+r)$, the ratio of the upper degeneracy to the total,
so a level scheme with $g_1 > g_0$ ends with more population above than below even
though the upper level costs energy, the entropy of the extra upper states paying
for the energy. The entropy at high temperature approaches $k_B\ln(g_0 + g_1)$
rather than $k_B\ln 2$, the logarithm of the total number of states, which is the
degeneracy showing up as extra accessible microstates. Take a lower singlet
$g_0 = 1$ and an upper triplet $g_1 = 3$, the pattern of a spin-one paramagnetic
ion with a singlet ground state. The Schottky heat-capacity peak shifts because the
effective gap in the occupation is $\epsilon - k_B T\ln r$: the degeneracy acts like
an entropic lowering of the gap by $k_B T\ln 3$, so the peak moves to a lower
temperature and grows taller than the non-degenerate Schottky anomaly of
figure~\ref{fig:vol04ch04:two-level-system}. Numerically the peak of the $r = 3$
system sits near $k_B T \approx 0.33\,\epsilon$ rather than the $0.42\,\epsilon$ of
the non-degenerate case, and its high-temperature entropy ceiling is $k_B\ln 4 =
1.39\,k_B$ against $k_B\ln 2 = 0.69\,k_B$. The lesson generalises to any level
scheme: the entropy a system can store is the logarithm of its total state count,
degeneracies included, and the temperature at which it stores heat most readily is
set by the energy gap offset by the entropic pull of the degeneracy ratio. A
magnetic refrigerant is chosen for a large ground-state degeneracy precisely because
that large $g$ gives a large spin entropy to unload in the demagnetization stroke of
the earlier example.

\subsection{The Gibbs phase rule and the degrees of freedom of a mixture}

Equilibrium fixes some of a system's intensive variables and leaves others free,
and the Gibbs phase rule counts exactly how many remain free from the chemical-
potential balances of section~4.6. We derive the rule from those balances and apply
it to read the degrees of freedom off a water phase diagram and a two-component
alloy.

A system of $C$ independent components distributed among $P$ phases in equilibrium
is described by the temperature, the pressure, and the composition of each phase.
Each phase's composition is set by $C - 1$ independent mole fractions, the last
fixed by summing to one, so the intensive state carries $2 + P(C-1)$ variables in
all. Equilibrium imposes constraints: the chemical potential of every component must
be equal across every phase, which for each of the $C$ components gives $P - 1$
independent equalities, $C(P-1)$ constraints in total. The number of intensive
variables free to vary, the degrees of freedom $F$, is the count of variables less
the count of constraints,
\[
F = 2 + P(C-1) - C(P-1) = C - P + 2,
\]
the Gibbs phase rule. For pure water, $C = 1$, so $F = 3 - P$. A single phase, liquid
water alone, has $F = 2$: temperature and pressure vary independently, which is why
the liquid region of the phase diagram is an area. Two phases in equilibrium, liquid
and vapour along the boiling curve, have $F = 1$: fixing the temperature fixes the
pressure, which is why coexistence is a line, the Clausius-Clapeyron curve of the
earlier derivation. Three phases, the triple point where solid, liquid, and vapour
meet, have $F = 0$: the state is fixed with no freedom at all, which is why the
triple point is a single point and why it serves as a temperature fixed point. For a
two-component alloy, $C = 2$ and $F = 4 - P$; a single-phase melt has three degrees
of freedom, a two-phase region where solid and liquid coexist has two, and the
eutectic, where two solids and a liquid coexist at once, has one at fixed pressure,
which is why a eutectic alloy melts at a single sharp temperature like a pure
substance. The phase rule is a bookkeeping consequence of the chemical-potential
equalities, and it tells an engineer reading any phase diagram how many knobs remain
to turn once a set of phases is required to coexist, before a single further
adjustment forces a phase to appear or vanish.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-kelvin-scale.tex}

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-solar-cell-limit.tex}

\section{Project}\pathtag{core}

\begin{project}[Simulate a small ideal gas and measure spontaneous
entropy increase]
The reader writes a simulation of a few hundred hard-sphere or
Lennard-Jones particles in a box, initialises them with all particles
in one half, runs the dynamics, and measures the time evolution of
the entropy.

\textbf{Track}. Simulation plus analysis.
\textbf{Hazard class}: none.

\textbf{Simulation}.

\begin{itemize}[leftmargin=*]
  \item Implement a two-dimensional (or three-dimensional)
    molecular-dynamics simulation of $N$ hard-disk (or Lennard-Jones)
    particles in a square (or cubic) box with periodic boundary
    conditions. Use $N$ in the range $100$ to $1000$.
  \item Initialise all particles in the left half of the box, with
    random velocities drawn from a Maxwell-Boltzmann distribution at
    a chosen temperature.
  \item Run the simulation forward in time. Record the number of
    particles in the left half at regular time intervals.
\end{itemize}

\textbf{Analysis}.

\begin{itemize}[leftmargin=*]
  \item Plot the number of particles in the left half versus time.
    Observe the approach to equipartition.
  \item Compute the coarse-grained entropy from the distribution at
    each time, using the formula $S = -k_B \sum_b p_b \ln p_b$ where
    $p_b$ is the fraction of particles in spatial bin $b$.
  \item Plot the entropy versus time. Identify the relaxation time
    to equilibrium and compare to the mean free time.
  \item Run the simulation in reverse from the equilibrated state
    (by reversing all velocities) and observe whether the system
    returns to the initial low-entropy configuration. Discuss the
    time-reversal-symmetry implications.
  \item Repeat for several values of $N$ and demonstrate the
    qualitative scaling of the fluctuations with $N$.
\end{itemize}

\textbf{Reflection ($1{,}500$ to $2{,}500$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
  \item How well does the simulated entropy obey the second law? At
    what particle count $N$ do fluctuations become noticeable?
  \item What is the role of the coarse-graining (the bin width) in
    determining the measured entropy?
  \item Under time-reversal of the velocities, does the system
    return to its initial low-entropy state? What does this say
    about the second law's status?
\end{itemize}

A reference implementation lives in \texttt{code/ideal\_gas\_md.py}: a
two-dimensional hard-disk integrator that initialises the disks in the
left half, draws velocities from a Maxwell-Boltzmann distribution, and
records the left-half fraction and the coarse-grained entropy over time.
Its output is \texttt{data/entropy\_relaxation.csv}, plotted in
figure~\ref{fig:vol04ch04:entropy-relaxation}: the entropy rises from a
low initial value toward $\ln 2$ per particle and then fluctuates around
the equilibrium value, with the fluctuation amplitude scaling as
$N^{-1/2}$. The reader is expected to extend the reference code, not
merely run it.

\input{volumes/04-energy/ch04-entropy-and-arrow-of-time/figures/fig-entropy-relaxation.tex}

\textbf{Deliverable}. The simulation code, the entropy-versus-time
plots, the time-reversal experiment, and the reflection.

\textbf{Time}. Eight to sixteen hours of programming and simulation,
four to six hours of write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Compute the entropy change of $10\,\text{g}$ of ice melting at
$0\,\degC$. The latent heat of fusion is $334\,\text{kJ}/\text{kg}$.

\paragraph{Solution sketch.} Melting at the fixed phase-change
temperature is reversible isothermal heat absorption, so
$\Delta S = Q/T$ with $Q = mL = 0.010 \times 334{,}000 = 3340\,\text{J}$
and $T = 273.15\,\si{\kelvin}$. Then
$\Delta S = 3340/273.15 = 12.2\,\si{\joule\per\kelvin}$, positive
because heat enters the system. The surroundings supplying the heat at
the same temperature lose $12.2\,\si{\joule\per\kelvin}$, so the total
entropy change is zero in the reversible limit; any finite temperature
gap between the reservoir and the ice makes the total positive.
\end{exercise}

\begin{exercise}
A system has two microstates with energies $0$ and $\epsilon$. At
temperature $T$, compute the partition function, the probability of
each state, the average energy, and the entropy.

\paragraph{Solution sketch.} The partition function is
$Z = 1 + e^{-\epsilon/k_B T}$. The probabilities are
$p_0 = 1/Z$ and $p_1 = e^{-\epsilon/k_B T}/Z = 1/(1 + e^{\epsilon/k_B T})$.
The average energy is
$\langle E \rangle = \epsilon p_1 = \epsilon/(1 + e^{\epsilon/k_B T})$.
The entropy follows from $S = (\langle E \rangle - F)/T$ with
$F = -k_B T \ln Z$, giving
$S = -k_B(p_0 \ln p_0 + p_1 \ln p_1)$. At $T \to 0$ the system sits in
the ground state ($p_0 \to 1$, $S \to 0$); at $T \to \infty$ the two
states are equally populated ($p_0 = p_1 = \tfrac12$,
$S \to k_B \ln 2$). The closed forms are checked numerically in
\texttt{code/partition\_function.py}.
\end{exercise}

\begin{exercise}
Compute the mixing entropy when $1\,\text{mol}$ of nitrogen and
$2\,\text{mol}$ of oxygen at the same temperature and pressure are
mixed to form an ideal-gas mixture.

\paragraph{Solution sketch.} For distinguishable ideal gases,
$\Delta S_{\text{mix}} = -R \sum_i n_i \ln x_i$, with mole fractions
$x_{N_2} = 1/3$ and $x_{O_2} = 2/3$. Then
$\Delta S_{\text{mix}} = -R[\,1 \cdot \ln(1/3) + 2 \cdot \ln(2/3)\,]
= -8.314\,[\,(-1.099) + 2(-0.405)\,]
= -8.314 \times (-1.910) = 15.9\,\si{\joule\per\kelvin}$.
The result is positive, as mixing distinguishable species always
increases entropy. Were the two gases identical, the Gibbs
indistinguishability correction would cancel the term and the mixing
entropy would be zero; figure~\ref{fig:vol04ch04:entropy-of-mixing}
plots the per-mole curve.
\end{exercise}

\begin{exercise}
For the water-gas-shift reaction with $\Delta H^\circ = -41.2\,
\text{kJ}/\text{mol}$ and $\Delta S^\circ = -42.1\,\text{J}/(\text{mol}
\cdot\si{\kelvin})$, compute the equilibrium constant at $25\,\degC$
and at $400\,\degC$. State the direction of the spontaneous reaction
at each temperature.

\paragraph{Solution sketch.} Use $\Delta G^\circ = \Delta H^\circ -
T\Delta S^\circ$ and $K_{\text{eq}} = \exp(-\Delta G^\circ/RT)$.
At $T = 298\,\si{\kelvin}$:
$\Delta G^\circ = -41{,}200 - 298(-42.1) = -41{,}200 + 12{,}546 =
-28{,}654\,\text{J}/\text{mol}$, so
$K_{\text{eq}} = \exp(28{,}654/(8.314 \times 298)) \approx 1.0 \times
10^{5}$; the forward reaction is strongly favoured.
At $T = 673\,\si{\kelvin}$:
$\Delta G^\circ = -41{,}200 + 673 \times 42.1 = -41{,}200 + 28{,}333 =
-12{,}867\,\text{J}/\text{mol}$, so
$K_{\text{eq}} = \exp(12{,}867/(8.314 \times 673)) \approx 10$; the
forward direction is still favoured but far less so. The crossover to
the reverse direction occurs near
$T^\ast = \Delta H^\circ/\Delta S^\circ \approx 978\,\si{\kelvin}$.
The sweep is in \texttt{data/wgs\_equilibrium.csv}.
\end{exercise}

\begin{exercise}
Compute the Landauer minimum energy for erasing $10^{10}$ bits at
room temperature ($T = 300\,\si{\kelvin}$). Compare to the energy
consumption of a typical computer for similar throughput.

\paragraph{Solution sketch.} The Landauer bound per bit is
$k_B T \ln 2 = 1.381 \times 10^{-23} \times 300 \times 0.693 =
2.87 \times 10^{-21}\,\text{J}$. For $10^{10}$ bits the minimum is
$10^{10} \times 2.87 \times 10^{-21} = 2.9 \times 10^{-11}\,\text{J}$,
about $30\,\text{pJ}$. A processor performing $10^{10}$ bit operations
dissipates on the order of joules (nanojoules per operation in current
hardware), so real computation runs roughly $10^{8}$ to $10^{11}$ times
above the Landauer floor. The bound is not the binding constraint on
today's chips; switching energy, leakage, and interconnect dominate.
The gap is the headroom that reversible and adiabatic logic chase.
\end{exercise}

\begin{exercise}
Estimate the entropy decrease of a cup of hot tea cooling from
$80\,\degC$ to $20\,\degC$ (mass $250\,\text{g}$, $c = 4.18\,\text{kJ}
/(\text{kg}\cdot\si{\kelvin})$) and the entropy increase of the
$20\,\degC$ surroundings. Verify the second law on the total.

\paragraph{Solution sketch.} The tea cools over a temperature range,
so integrate: $\Delta S_{\text{tea}} = mc \ln(T_f/T_i) =
0.250 \times 4180 \times \ln(293/353) = 1045 \times (-0.186) =
-194\,\si{\joule\per\kelvin}$. The heat shed is
$Q = mc\,\Delta T = 1045 \times 60 = 62{,}700\,\text{J}$, absorbed by
the surroundings at the fixed $293\,\si{\kelvin}$:
$\Delta S_{\text{surr}} = +62{,}700/293 = +214\,\si{\joule\per\kelvin}$.
The total is $-194 + 214 = +20\,\si{\joule\per\kelvin} > 0$, satisfying
the second law. The positive total is the entropy generated by heat
crossing the finite temperature gap; it vanishes only if the cooling is
carried out reversibly through a sequence of reservoirs.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
From the canonical partition function $Z = \sum_i e^{-E_i/k_B T}$,
derive the relations $\langle E \rangle = -\partial \ln Z/\partial
\beta$ and $F = -k_B T \ln Z$.

\paragraph{Solution sketch.} Write $Z = \sum_i e^{-\beta E_i}$ with
$\beta = 1/k_B T$. Differentiate:
$\partial Z/\partial \beta = -\sum_i E_i e^{-\beta E_i}$, so
$-\partial \ln Z/\partial \beta = (1/Z)\sum_i E_i e^{-\beta E_i}
= \sum_i E_i p_i = \langle E \rangle$, which is the first relation.
For the free energy, start from the statistical entropy
$S = -k_B \sum_i p_i \ln p_i$ with $p_i = e^{-\beta E_i}/Z$. Then
$\ln p_i = -\beta E_i - \ln Z$, so
$S = -k_B \sum_i p_i(-\beta E_i - \ln Z)
= k_B \beta \langle E \rangle + k_B \ln Z
= \langle E \rangle/T + k_B \ln Z$. Rearranging,
$\langle E \rangle - TS = -k_B T \ln Z$, and the left side is the
Helmholtz free energy $F$, giving $F = -k_B T \ln Z$.
\end{exercise}

\begin{exercise}
Derive the Boltzmann distribution $P_i \propto e^{-E_i/k_B T}$ by
maximising the Gibbs entropy $S = -k_B \sum_i p_i \ln p_i$ subject to
the constraints of normalisation and fixed average energy.

\paragraph{Solution sketch.} Maximise $S/k_B = -\sum_i p_i \ln p_i$
with two Lagrange multipliers, $\alpha$ for normalisation
$\sum_i p_i = 1$ and $\beta$ for the energy $\sum_i p_i E_i =
\langle E \rangle$. Form
$\mathcal{L} = -\sum_i p_i \ln p_i - \alpha(\sum_i p_i - 1)
- \beta(\sum_i p_i E_i - \langle E \rangle)$ and set
$\partial \mathcal{L}/\partial p_i = -\ln p_i - 1 - \alpha
- \beta E_i = 0$. Solving gives
$p_i = e^{-1-\alpha} e^{-\beta E_i}$, that is
$p_i \propto e^{-\beta E_i}$. Normalisation fixes the prefactor to
$1/Z$ with $Z = \sum_i e^{-\beta E_i}$. Identifying the multiplier
$\beta$ with $1/k_B T$ (by matching $\partial S/\partial \langle E
\rangle = 1/T$) recovers the Boltzmann distribution. The second-order
condition confirms a maximum, since $-\sum_i p_i \ln p_i$ is concave in
the $p_i$.
\end{exercise}

\begin{exercise}
Derive the equipartition theorem: each quadratic degree of freedom in
the Hamiltonian contributes $\frac{1}{2} k_B T$ to the average
energy.

\paragraph{Solution sketch.} Take one quadratic coordinate, $E = a q^2$
with $a > 0$, in the classical canonical ensemble. Its average energy is
\[
\langle E \rangle = \frac{\int_{-\infty}^{\infty} a q^2
e^{-\beta a q^2}\,dq}{\int_{-\infty}^{\infty} e^{-\beta a q^2}\,dq}.
\]
Use $\langle E \rangle = -\partial \ln Z_q/\partial \beta$ with
$Z_q = \int e^{-\beta a q^2}\,dq = \sqrt{\pi/(\beta a)}
\propto \beta^{-1/2}$. Then $\ln Z_q = -\tfrac12 \ln \beta +
\text{const}$ and $-\partial \ln Z_q/\partial \beta = 1/(2\beta) =
\tfrac12 k_B T$. Each independent quadratic term (each squared momentum
or squared displacement in the Hamiltonian) contributes the same
$\tfrac12 k_B T$, so a monatomic gas with three translational squares
has $\langle E \rangle = \tfrac32 k_B T$ per molecule. The derivation is
classical; it fails when $k_B T$ drops below a mode's quantum spacing,
which is why rotational and vibrational modes freeze out at low
temperature.
\end{exercise}

\begin{exercise}
Show that for an ideal monatomic gas, the canonical partition
function gives the Sackur-Tetrode entropy
\[
S = N k_B \left[\ln \left(\frac{V}{N \Lambda^3}\right) +
\frac{5}{2}\right],
\]
where $\Lambda$ is the thermal de Broglie wavelength.

\paragraph{Solution sketch.} The single-molecule translational
partition function is $z_1 = V/\Lambda^3$ with
$\Lambda = h/\sqrt{2\pi m k_B T}$. With the Gibbs factor,
$Z = z_1^N/N! = (V/\Lambda^3)^N/N!$. Then
$F = -k_B T \ln Z = -k_B T[\,N \ln(V/\Lambda^3) - \ln N!\,]$. Apply
Stirling, $\ln N! \approx N \ln N - N$, to get
$F = -N k_B T[\,\ln(V/N\Lambda^3) + 1\,]$. The entropy is
$S = -(\partial F/\partial T)_V$. The temperature enters through
$\Lambda \propto T^{-1/2}$, so $\ln(V/N\Lambda^3) = \ln(V/N) +
\tfrac32 \ln T + \text{const}$ and
$\partial F/\partial T$ carries the explicit $T$ in front plus the
$\tfrac32$ from the $\Lambda$ dependence. Collecting terms gives
$S = N k_B[\,\ln(V/N\Lambda^3) + \tfrac52\,]$, the Sackur-Tetrode
result. The $\tfrac52$ is $1$ from the additive constant plus
$\tfrac32$ from the translational temperature dependence.
\end{exercise}

\begin{exercise}
Derive the van't Hoff equation $d \ln K_{\text{eq}}/dT = \Delta
H^\circ/(RT^2)$ from the temperature dependence of $\Delta G^\circ =
-RT \ln K_{\text{eq}}$.

\paragraph{Solution sketch.} From $\Delta G^\circ = -RT \ln
K_{\text{eq}}$, write $\ln K_{\text{eq}} = -\Delta G^\circ/(RT)$.
Use the Gibbs-Helmholtz relation
$\partial(\Delta G^\circ/T)/\partial T = -\Delta H^\circ/T^2$, which
follows from $\Delta G^\circ = \Delta H^\circ - T\Delta S^\circ$ by
differentiating $\Delta G^\circ/T = \Delta H^\circ/T - \Delta S^\circ$
at constant $\Delta H^\circ, \Delta S^\circ$. Then
$\frac{d \ln K_{\text{eq}}}{dT} = -\frac{1}{R}
\frac{d(\Delta G^\circ/T)}{dT}
= -\frac{1}{R}\left(-\frac{\Delta H^\circ}{T^2}\right)
= \frac{\Delta H^\circ}{RT^2}$.
The sign of $\Delta H^\circ$ sets the direction: $K_{\text{eq}}$ rises
with temperature for endothermic reactions and falls for exothermic
ones, the temperature form of Le Chatelier's principle.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the entropy of a deck of cards in a well-shuffled state
relative to the perfectly ordered state. Express the answer in bits
and in $\text{J}/\si{\kelvin}$.

\paragraph{Reference estimate.} A shuffled deck can be in any of
$52!$ orderings, the perfectly ordered deck in one. The information
entropy is $H = \log_2(52!) \approx 225.6$ bits, using
$\log_2(52!) = \log_2(8.07 \times 10^{67}) = 67.9/0.301 \approx 226$.
Converting to thermodynamic units, $S = k_B \ln(52!) =
1.381 \times 10^{-23} \times \ln(8.07 \times 10^{67})
= 1.381 \times 10^{-23} \times 156.4 = 2.2 \times 10^{-21}\,
\si{\joule\per\kelvin}$. The thermodynamic entropy of the ordering is
vanishingly small next to the $\sim 10^2\,\si{\joule\per\kelvin}$
thermal entropy of the cardboard itself. The point of the estimate is
that information entropy and thermodynamic entropy are the same quantity
in different units, related by the factor $k_B \ln 2$ per bit.
\end{exercise}

\begin{exercise}
Estimate the number of microstates corresponding to one mole of an
ideal gas at standard conditions. State the assumed cell size in
phase space.

\paragraph{Reference estimate.} Take the absolute entropy of one mole
of a simple gas from the Sackur-Tetrode formula, of order $S \approx
150\,\si{\joule\per\kelvin}$ (for argon, $S^\circ = 155\,
\si{\joule\per\kelvin}$ at standard conditions). Invert
$S = k_B \ln W$ to get
$\ln W = S/k_B = 150/(1.381 \times 10^{-23}) = 1.09 \times 10^{25}$,
so $W = e^{1.09 \times 10^{25}} = 10^{(1.09 \times 10^{25})/2.303}
\approx 10^{4.7 \times 10^{24}}$. The phase-space cell size is set by
the quantum of action $h^{3N}$ for $N = N_A$ particles, which is the
choice implicit in the Sackur-Tetrode entropy through the thermal de
Broglie wavelength $\Lambda^3 \propto h^3$. Without that quantum cell
the count is ambiguous by an additive constant; with it, $W$ is the
absolute microstate number, of order $10^{10^{25}}$ in rough form.
\end{exercise}

\begin{exercise}
Estimate the rate of entropy production by a $60\,\text{W}$
incandescent light bulb radiating in a $20\,\degC$ room. State the
filament temperature assumed.

\paragraph{Reference estimate.} In steady state the bulb converts
$60\,\text{W}$ of electrical work, which carries no entropy, into heat
and light that the room ultimately absorbs at $T_0 = 293\,\si{\kelvin}$.
The entropy leaves the filament (at $T_f \approx 2800\,\si{\kelvin}$)
at rate $\dot S_{\text{out}} = 60/2800 = 0.021\,\text{W}/\si{\kelvin}$
and arrives in the room at $\dot S_{\text{in}} = 60/293 =
0.205\,\text{W}/\si{\kelvin}$. The net entropy generation is
$\dot S_{\text{gen}} = 60(1/293 - 1/2800) =
0.205 - 0.021 = 0.18\,\text{W}/\si{\kelvin}$. The dominant generation is
the degradation of high-temperature radiation and electrical work into
room-temperature heat. The bulb is an almost perfect entropy generator:
nearly all the supplied work ends as low-grade heat.
\end{exercise}

\begin{exercise}
Estimate the entropy increase associated with respiration: an adult
human consumes about $100\,\text{W}$ of metabolic power, rejecting
the heat at body temperature ($310\,\si{\kelvin}$) to a $295\,\si{
\kelvin}$ environment. Compute the rate of entropy export to the
environment.

\paragraph{Reference estimate.} The body takes in $100\,\text{W}$ of
chemical free energy and, in steady state, rejects $100\,\text{W}$ as
heat. The heat enters the body's tissues and is shed to the environment
at $T_{\text{env}} = 295\,\si{\kelvin}$, so the entropy exported to the
environment is $\dot S_{\text{env}} = 100/295 =
0.34\,\text{W}/\si{\kelvin}$. The entropy carried by the incoming
chemical fuel is far smaller (food is low-entropy, high-free-energy
matter), so the body is a net entropy exporter at roughly
$0.34\,\text{W}/\si{\kelvin}$, the rate at which it maintains its own
internal order. Over a day this is about
$0.34 \times 86{,}400 = 2.9 \times 10^4\,\si{\joule\per\kelvin}$ shed to
the surroundings.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Simulate $N$ particles in a box and measure the relaxation time to
equipartition starting from a half-box initial condition. Vary $N$
from $10$ to $10^4$ and measure how the fluctuation amplitude scales
with $N$.

\paragraph{Solution sketch.} Use the reference code
\texttt{code/ideal\_gas\_md.py}, which starts the disks in the left
half and records the left-half fraction and coarse-grained entropy. The
relaxation time, measured as the time for the left-half fraction to
reach $1/2$ within a tolerance, is a few mean free times and is nearly
independent of $N$ once $N$ is large. The equilibrium fluctuation in the
left-half count has standard deviation $\sqrt{N}/2$ in absolute terms,
so the relative fluctuation in the fraction scales as $N^{-1/2}$: a
factor $10^{1.5} \approx 32$ smaller going from $N = 10$ to $N = 10^4$.
Plotting the root-mean-square fluctuation of the equilibrated fraction
against $N$ on log-log axes should give a slope of $-1/2$;
figure~\ref{fig:vol04ch04:entropy-relaxation} shows the qualitative
difference between $N = 20$ and $N = 1000$.
\end{exercise}

\begin{exercise}
Implement a Metropolis Monte Carlo simulation of an Ising model with
$N$ spins and compute the average energy and heat capacity as
functions of temperature. Identify the critical temperature and
discuss the entropy near the transition.

\paragraph{Solution sketch.} For a two-dimensional square-lattice
Ising model with nearest-neighbour coupling $J$, the Metropolis rule
flips a spin with probability $\min(1, e^{-\Delta E/k_B T})$. Sweep the
temperature and record $\langle E \rangle$ and the heat capacity, either
as $C = (\langle E^2\rangle - \langle E\rangle^2)/(k_B T^2)$ from the
fluctuation formula or as a numerical $dE/dT$. The heat capacity peaks
sharply at the Onsager critical temperature
$k_B T_c/J = 2/\ln(1+\sqrt2) \approx 2.269$. Near $T_c$ the entropy
rises steeply as the spins disorder; the heat-capacity peak is the
signature of the second-order transition, where correlations diverge
and the system passes from the ordered (low-entropy, aligned) phase to
the disordered (high-entropy) phase. Finite $N$ rounds the peak;
extrapolating the peak position with system size recovers the
thermodynamic $T_c$.
\end{exercise}

\begin{exercise}
Use a stochastic-trajectory simulation to verify the Jarzynski
equality $\langle e^{-W/k_B T} \rangle = e^{-\Delta F/k_B T}$ for a
simple system with a known free-energy difference.

\paragraph{Solution sketch.} The reference code
\texttt{code/jarzynski\_demo.py} drags an overdamped particle in a
harmonic trap whose centre moves at constant speed. Because the trap
stiffness is held fixed, the free-energy difference is zero, so the
target is $\langle e^{-W/k_B T}\rangle = 1$. Each realisation gives a
work value; the mean work is positive (the pull dissipates energy,
$\langle W\rangle \ge \Delta F$), yet the exponential average of the
work returns one to within sampling error over $\sim 2\times10^4$
realisations. The exponential average is dominated by rare trajectories
with small or negative work, which is why the estimator needs many
samples and converges slowly when the dissipation is large. The
trajectory-level negative-work events are exactly the fluctuation
content the failure section describes: forbidden on average, allowed and
necessary in individual runs.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify the cooling load and the minimum compressor work for an
industrial freezer that holds $10\,\text{tonnes}$ of product at
$-30\,\degC$ in a $25\,\degC$ ambient. State the heat-leak rate and
the second-law efficiency assumed.

\paragraph{Model-answer sketch.} Assume the steady-state load is the
heat leak through the enclosure (the product is already cold, so the
pull-down load is transient). Take a well-insulated cold room with
overall conductance giving a leak of order $5\,\text{kW}$ across the
$55\,\si{\kelvin}$ gradient (state this assumption explicitly; it
follows from an envelope of $\sim 200\,\text{m}^2$ at
$U \approx 0.4\,\text{W}/(\text{m}^2\,\si{\kelvin})$). The Carnot
coefficient of performance for cooling is
$\mathrm{COP}_{\text{Carnot}} = T_C/(T_H - T_C) = 243/(298 - 243) =
4.4$. The reversible compressor work is
$\dot W_{\min} = \dot Q_C/\mathrm{COP}_{\text{Carnot}} = 5000/4.4 =
1.14\,\text{kW}$. A real plant runs at a second-law efficiency of about
$0.4$, so the installed compressor power is
$\dot W_{\text{real}} = 1.14/0.4 \approx 2.8\,\text{kW}$. The product
mass enters only the pull-down: cooling $10\,\text{tonnes}$ of, say,
water-like product by $55\,\si{\kelvin}$ removes
$10{,}000 \times 4180 \times 55 = 2.3\,\text{GJ}$, which at the leak
power would take many hours, so the pull-down sizes the compressor if a
fast pull-down is required.
\end{exercise}

\begin{exercise}
Design a thermoelectric cooler to maintain a single $1\,\text{cm}^3$
sample at $-10\,\degC$ in a $25\,\degC$ ambient. State the
thermoelectric figure of merit $ZT$ assumed and the resulting cooling
power.

\paragraph{Model-answer sketch.} The gradient is
$\Delta T = 35\,\si{\kelvin}$ with $T_C = 263\,\si{\kelvin}$,
$T_H = 298\,\si{\kelvin}$. A commercial bismuth-telluride module has
$ZT \approx 0.7$ at room temperature (state this). The maximum coefficient
of performance of a thermoelectric cooler is
$\mathrm{COP}_{\max} = \frac{T_C}{\Delta T}
\frac{\sqrt{1+ZT_m}-T_H/T_C}{\sqrt{1+ZT_m}+1}$, with
$T_m$ the mean temperature. With $\sqrt{1+ZT_m} = \sqrt{1.7} = 1.30$,
the second factor is $(1.30 - 1.133)/(1.30 + 1) = 0.073$, so
$\mathrm{COP}_{\max} = (263/35)\times 0.073 \approx 0.55$. A
$1\,\text{cm}^3$ sample with a modest heat leak of $\sim 0.5\,\text{W}$
then needs about $0.5/0.55 \approx 0.9\,\text{W}$ of electrical input.
The low $ZT$ is why thermoelectric cooling suits small loads and loses
to vapour-compression at large scale: its second-law efficiency is far
below that of a compressor cycle. The cooling power is set by the leak,
the achievable $\Delta T$ by the module's $ZT$.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A small reaction calorimeter measures an apparent entropy decrease in
an isolated sample of $-0.1\,\text{J}/\si{\kelvin}$ over $1\,\text{h}$.
Identify the most probable source of measurement error and the
diagnostic measurement.

\paragraph{Solution sketch.} An isolated sample cannot decrease its own
total entropy, so the boundary is wrong: the sample is not actually
isolated. The most probable error is an unaccounted heat leak out of the
calorimeter, which carries entropy to the surroundings and looks like an
internal decrease when only the sample is booked. A leak of order
$0.1\,\text{J}/\si{\kelvin}\times 300\,\si{\kelvin} \approx 30\,\text{J}$
of heat over the hour, about $8\,\text{mW}$, is well within the leak rate
of an ordinary small calorimeter. The diagnostic is a blank run with no
reaction: if the empty cell shows the same apparent drift, the
calorimeter has a baseline heat leak or a sensor drift, not a real
entropy change. Improve the insulation, add a guard heater, or subtract
the measured baseline.
\end{exercise}

\begin{exercise}
A protein-folding experiment finds that a denatured protein
spontaneously refolds at body temperature, with an estimated entropy
change of $-150\,\text{J}/(\text{mol}\cdot\si{\kelvin})$. Identify
the source of the entropy that is exported to the environment and
discuss whether the process is consistent with the second law.

\paragraph{Solution sketch.} The protein's configurational entropy
falls as it folds (the chain loses access to its many denatured
conformations), $\Delta S_{\text{sys}} = -150\,\text{J}/(\text{mol}\,
\si{\kelvin})$. Folding is spontaneous at body temperature, so
$\Delta G < 0$, which requires the enthalpy term to dominate:
$\Delta G = \Delta H - T\Delta S_{\text{sys}} < 0$ with
$T\Delta S_{\text{sys}} = 310\times(-150) = -46.5\,\text{kJ}/\text{mol}$,
so a folding enthalpy more negative than about
$-46.5\,\text{kJ}/\text{mol}$ (favourable hydrogen bonds, van der Waals
contacts, hydrophobic burial) makes $\Delta G < 0$. That enthalpy is
released as heat to the surrounding water, raising the environment's
entropy by $|\Delta H|/T \ge 150\,\text{J}/(\text{mol}\,\si{\kelvin})$,
which at least offsets the protein's entropy loss. The hydrophobic
effect also contributes: releasing ordered water from around exposed
residues raises the solvent entropy. The process is consistent with the
second law because $\Delta S_{\text{total}} = \Delta S_{\text{sys}} +
\Delta S_{\text{surr}} \ge 0$; the system orders by exporting entropy to
the solvent.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A 2024 venture pitch claims a ``microthermal energy harvester'' that
extracts $1\,\text{W}$ of electrical work per square centimetre from
ambient temperature fluctuations of $\pm 0.1\,\si{\kelvin}$. Apply
the four-step diagnostic of section 4.7 to identify the violation.

\paragraph{Solution sketch.} \emph{Redraw the boundary}: a single
ambient reservoir at one temperature is the only declared source; there
is no second reservoir at a different temperature, so there is no
sustained gradient to draw work from. \emph{Recount the flows}: a
fluctuation of $\pm 0.1\,\si{\kelvin}$ is a temporal wobble of one
reservoir, not a standing temperature difference between two bodies;
the Kelvin-Planck statement forbids extracting net work from a single
reservoir in a cycle. \emph{Recompute the bound}: even treating the
fluctuation as a transient $\Delta T = 0.1\,\si{\kelvin}$, the Carnot
factor is $\Delta T/T \approx 0.1/300 = 3\times10^{-4}$; harvesting
$1\,\text{W}$ of work would require a heat throughput of
$1/(3\times10^{-4}) \approx 3\,\text{kW}$ through one square centimetre,
which the thermal mass and conductance of any real device cannot
sustain. \emph{Replicate}: the device, run against a single
well-stirred reservoir, will produce zero net work. The claim is a
single-reservoir engine, the canonical second-law violation; the
fluctuations are real but carry no exploitable free energy without a
second reservoir at a different temperature.
\end{exercise}

\begin{exercise}
A peer-reviewed paper reports observing a $10\,\%$ decrease in
entropy in a closed isolated system over a measurement period of
$1\,\text{hour}$. Identify the most probable explanations from among
(a) the system was not isolated, (b) the entropy formula was applied
incorrectly, (c) the measurement was a statistical fluctuation, and
state the diagnostic that distinguishes them.

\paragraph{Solution sketch.} Rank by prior probability. \emph{(a) Not
isolated} is most likely: a $10\,\%$ entropy decrease over an hour
implies a large, slow heat or matter leak that the boundary did not
book; the diagnostic is a blank run and an independent energy balance
across every boundary. \emph{(b) Formula misapplied} is next: entropy of
an open or non-equilibrium system computed with an equilibrium formula,
or a coarse-graining changed mid-measurement, can manufacture a spurious
decrease; the diagnostic is to recompute $S$ from first principles with
a fixed, stated set of macroscopic variables. \emph{(c) Statistical
fluctuation} is effectively impossible for a macroscopic closed system:
a $10\,\%$ decrease corresponds to a probability of order
$e^{-\Delta S/k_B}$ with $\Delta S$ of order $10^{22}\,k_B$, so the
expected wait far exceeds the age of the universe. The distinguishing
diagnostic is scale: a fluctuation of this size is allowed only in a
system of a handful of particles, where the fluctuation theorems apply;
for any macroscopic sample, (a) or (b) is the explanation.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A philosopher argues that the second law is a tautology because
``entropy is defined to increase.'' Write a one-paragraph reply that
distinguishes the operational definition of entropy from the
empirical content of the second law.

\paragraph{Model-answer sketch.} Entropy is defined operationally by
measurement, $dS = \delta Q_{\text{rev}}/T$ along a reversible path, or
microscopically by $S = k_B \ln W$ from a count of states; neither
definition mentions a direction of time. The empirical content of the
second law is the separate, falsifiable claim that for an isolated
system this measured quantity never decreases. The claim could in
principle be wrong: a clear, reproducible isolated-system entropy
decrease would refute it, and two centuries of measurement have not
produced one. The law is not a tautology because the definition fixes
how to compute $S$ from data, while the law asserts a contingent fact
about how that computed $S$ behaves in time. The reply to the
philosopher is that ``entropy is defined to increase'' confuses the
measurement rule with the empirical regularity; the two are logically
independent.
\end{exercise}

\begin{exercise}
A junior engineer asks whether the second law is ``ever broken in
quantum systems.'' Write a one-paragraph reply that addresses the
fluctuation theorems and the operational status of the law at
nanoscale.

\paragraph{Model-answer sketch.} The second law holds as a statement
about averages, in quantum systems as in classical ones. At the level
of single trajectories of a small system, individual realisations can
show a transient negative entropy production; the Jarzynski and Crooks
relations make this quantitative, fixing the probability of a
negative-entropy trajectory relative to its positive-entropy mirror as
$e^{-\Delta S/k_B}$. For a nanoscale device with $\Delta S$ of order a
few $k_B$, these excursions are observable and have been measured; for
anything macroscopic the ratio is astronomically small and the law is
operationally absolute. So the honest answer is that the law is never
broken on average, the fluctuation theorems describe the trajectory-level
excursions exactly rather than as exceptions, and the engineering of
nanoscale machines uses those theorems as its working framework. Nothing
in quantum mechanics provides a route to a sustained, extractable
violation.
\end{exercise}
