\chapter{Regulatory and certification failure}
\label{vol10:ch11}

\epigraph{The certificate says the system meets the specification; the specification did not anticipate the failure mode that killed the people.}{}

\chapmeta{Half-life: durable. Archetypes: ethics, failure. Exercise target: 15. Project track: hybrid.}

\noindent
Regulation and certification are the institutional mechanisms by which engineering work is held accountable to the public's interest in safety. They produce documents (type certificates, operating licences, market approvals) and processes (design reviews, inspections, audits) that, when functioning, give the public reasonable assurance that systems entering service are safe to the standards that current engineering knowledge supports. When they fail, the failure mode is structural: not the individual inspector's lapse but the regulatory architecture's incapacity to keep pace with the technology, the industry, the political pressure, or its own staffing. This chapter is about those failures, the cases that exemplify them, and the obligations the engineer holds within the regulatory relationship.

\section{Regulators and engineering}\pathtag{core}

The regulator's function is to translate the public's interest in safety into requirements that the engineering profession can meet, into evidence the engineering work can produce, and into accountability the engineering organisations can be held to. The function divides into three sub-functions that are conceptually distinct and institutionally often separate:

\textbf{Standard-setting.} What level of safety must the system achieve; what test must it pass; what evidence must it produce; what design margins must it carry. The standard is the regulator's translation of public-interest claims into specifications the engineering can be measured against. Standards are sometimes set by the regulator alone (FDA's quality system regulations; NRC's Part 50 requirements), sometimes by independent standards bodies that the regulator adopts (ASTM standards adopted by OSHA; ISO standards referenced by national regulators; SAE standards in aerospace; ASME boiler and pressure vessel codes), sometimes by industry consensus that the regulator endorses.

\textbf{Pre-market evaluation.} Before the system enters service, is its design adequate against the standard; does the manufacturer's evidence support the claim of adequacy; is the manufacturer's manufacturing process sufficient to produce conforming articles. The pre-market evaluation produces the certificate, licence, or approval that authorises operation. The regulator may perform the evaluation directly (FDA pre-market approval for high-risk medical devices), may delegate the evaluation to authorised representatives (FAA Organisation Designation Authorisation for civil aircraft type certification), or may rely on a manufacturer's self-declaration of conformity (CE-marking under many European directives).

\textbf{Post-market surveillance.} After the system enters service, is it actually meeting the standard in field operation; have new failure modes emerged that the certification did not anticipate; are there manufacturing deviations, maintenance shortfalls, operational practices that warrant intervention. Post-market surveillance includes incident reporting (the FAA Service Difficulty Report system; the FDA's MAUDE database; the NRC's Licensee Event Reports), inspection (planned and event-driven), and enforcement action (warnings, mandates, certificate revocations).

The three functions interact. A standard that is set too loose makes the evaluation a formality; a standard that is set too tight produces evidence-gaming. An evaluation that is delegated to manufacturer-employed representatives risks the loyalty divisions identified in the 737 MAX case (Chapter 10) \cite{acc:house-737max-2020}. A surveillance system that depends on industry incident reporting risks the selective-reporting failure mode (Chapter 10) that production pressure produces. The functions also have different time constants: standards evolve slowly (years to decades), evaluations take months to years per programme, surveillance is ongoing. The mismatch in time constants is itself a source of failure: a standard adequate to last year's technology becomes inadequate to this year's; an evaluation conducted under one set of regulatory expectations becomes outdated when the expectations shift; surveillance data that takes years to accumulate may not be available when this year's certification decisions are made.

The engineer's relationship to the regulator depends on the function. For standard-setting, the engineer's role is technical contribution: participation in standards committees, in regulatory rulemaking comment periods, in industry-association technical activities. For pre-market evaluation, the engineer's role is evidence production: design documentation, test reports, analysis reports, hazard assessments that the regulator (or the delegated representative) reviews. For post-market surveillance, the engineer's role is honest reporting: incident reports, deviation notifications, root-cause analyses that the regulator can act on.

\section{Certification regimes: aviation, medical devices, nuclear, automotive}\pathtag{standard}

The dominant certification regimes are best understood by comparison; the differences between them illuminate the design choices and their consequences.

\textbf{Aviation type certification (FAA, EASA, CAAC, Transport Canada).} A new aircraft type requires a Type Certificate from the national authority (mutual recognition through bilateral agreements simplifies operation across jurisdictions). The certification process establishes that the design meets the airworthiness standards (FAR Part 23 for general aviation, Part 25 for transport-category aircraft, Part 27/29 for rotorcraft) through a combination of design review, ground test, and flight test. The process can take years for a major new type; for a major derivative (like the 737 MAX), the process may treat much of the underlying type design as already-certified and focus on the changes. The FAA's Organisation Designation Authorisation (ODA) delegates substantial certification work to manufacturer-employed engineers; the House Committee on 737 MAX identified the ODA arrangement as a contributing factor in the certification gaps \cite{acc:house-737max-2020}. Post-certification surveillance is through the Continued Airworthiness system: airworthiness directives mandate corrective action when a fleet-wide problem emerges; service difficulty reports flag field problems that may warrant such action.

\textbf{Medical device approval (FDA, EMA, MHRA, PMDA).} Medical devices in the US fall under three regulatory classes by risk: Class I (general controls; mostly exempt from pre-market notification), Class II (510(k) pre-market notification; substantial equivalence to a predicate device), Class III (PMA pre-market approval; clinical evidence of safety and effectiveness). The Therac-25 was regulated under the 510(k) pathway as substantially equivalent to predecessor radiation-therapy devices; the regulatory regime did not require independent review of the software architecture or independent test of the dose-control software \cite{paper:leveson-turner1993}. The post-Therac-25 regulatory evolution included the Safe Medical Devices Act of 1990 (which expanded reporting requirements), the FDA's general principles of software validation guidance (which became the basis for software-in-medical-devices oversight), and the eventual development of recognised consensus standards (IEC 62304 for medical-device software life-cycle processes; IEC 60601-1 for general electrical safety) that subsequent device approvals depend on.

\textbf{Nuclear plant licensing (NRC, ASN, ONR, CNSC).} Nuclear plants are licensed individually by the national regulator under a process that includes design certification (for the reactor type), construction permit (for the specific site), and operating licence. The licensing process is exhaustive, lasting years; the operating licence is conditional on continuing compliance with technical specifications, periodic inspections, and operator licensing. Post-Three-Mile-Island reforms (Chapter 9, \cite{acc:kemeny-tmi-1979}) included the establishment of the Institute of Nuclear Power Operations (INPO) as an industry peer-review body, the requirement for symptom-based emergency procedures, and the upgrade of operator training to full-scope simulator standards. Post-Chernobyl reforms (Chapter 10, \cite{acc:iaea-insag7-1992}) extended the framework internationally through the IAEA Convention on Nuclear Safety and the WANO peer-review system.

\textbf{Automotive (NHTSA, KBA, the Type Approval system in Europe, the SAE/ISO 26262 functional-safety standard).} Automotive regulation differs from aerospace in the relative weight of self-certification: in the US, manufacturers self-certify compliance with Federal Motor Vehicle Safety Standards (FMVSS), with the NHTSA's role primarily post-market (defect investigations, recalls). The European Type Approval system requires pre-market approval but the approval process is conducted by national authorities and there is substantial industry-led standard-setting through ISO and SAE. The Volkswagen diesel-emissions defeat device was implemented under a self-certification regime that did not require independent test of the engine-control software under the full range of operating conditions; the device's discovery came from independent on-road emissions testing conducted by ICCT and West Virginia University, not from regulatory test \cite{acc:epa-vw-nov-2015}.

The differences across regimes are instructive. Aviation is the most regulatory-intensive of the major engineering regimes, with detailed pre-market review and substantial post-market surveillance; its accident rate per passenger-mile is correspondingly low. Medical devices have a wider range of regulatory intensity, with high-risk devices receiving aviation-level review and lower-risk devices receiving minimal pre-market scrutiny. Nuclear is regulatory-intensive but small-volume (each plant is licensed individually); the regulatory cost per plant is high and is part of the industry's overall economics. Automotive has historically been the lightest-regulatory of the high-volume industries, with self-certification and post-market enforcement; this has been changing as software-defined vehicles raise the certification burden.

\section{Self-certification and its discontents}\pathtag{standard}

Self-certification is the regulatory arrangement in which the manufacturer attests to compliance with the applicable standards and the regulator's role is post-market enforcement of the attestation. The arrangement is attractive for its low pre-market burden (the regulator does not have to evaluate every product before market entry), but it shifts the assurance of compliance from regulator capability to manufacturer integrity.

The arrangements that produce credible self-certification are several. \textbf{Third-party conformity assessment}: an independent test laboratory, accredited to the relevant standards by a national accreditation body, conducts the tests and issues the conformity certificate. The arrangement is the basis of CE-marking under many European directives; the credibility depends on the accreditation body's discipline and the test laboratory's independence from the manufacturer. \textbf{Audited self-certification}: the manufacturer maintains the quality system and the conformity declarations; the regulator audits a sample of declarations and inspects the quality system. \textbf{Type-approval self-certification}: the manufacturer obtains a type approval (one-time, by a regulator or accredited body), then self-certifies that production conforms to the type. The arrangement is common in automotive Type Approval and in machinery directives.

The failure modes of self-certification:

\textbf{Conformance to specification but specification inadequate.} The product conforms to the standard; the standard does not anticipate the failure mode that occurs in the field. The post-failure response is to revise the standard, but the products already in the field were certified to the old standard and the response does not retroactively bring them into compliance.

\textbf{Test conditions different from field conditions.} The product passes the test; the field conditions differ from the test conditions; the product fails in the field in a way the test did not detect. The Volkswagen defeat device is the limiting case: the engine-control software detected the test condition and altered emissions behaviour for the duration of the test, producing certified-compliant emissions on the dynamometer and non-compliant emissions on the road \cite{acc:epa-vw-nov-2015}. The case is unusual in that the manufacturer deliberately exploited the test-condition gap; the more common case is that the gap is unintentional and the product's field behaviour is the genuine specification limit.

\textbf{Self-attestation without underlying compliance.} The manufacturer issues the conformity declaration without actually conducting the tests or maintaining the quality system the standard requires. The arrangement depends on enforcement: spot-check audits, market surveillance, post-incident investigation. The discovery rate is bounded by the regulator's surveillance capacity; in low-surveillance regimes, the discovery rate is low and the incentive to non-compliance is correspondingly high.

\textbf{Regulator-recognised standards body captured by industry.} The standard against which self-certification is measured is itself a product of an industry-led standards body whose members are the manufacturers who will be certified against the standard. The standard's stringency is bounded by what the industry can comfortably meet; the standard's evolution lags the technology's capability for substantially stricter performance. The arrangement is not formal regulator capture (the regulator does not itself decide the standard), but the effect on regulatory stringency is comparable.

The defences against self-certification failure are: post-market surveillance with meaningful enforcement (the recall mandates the regulator can issue; the criminal sanctions the regulator can refer); independent test capability (the regulator's own laboratories; the recognised independent laboratories whose certifications the regulator accepts); transparency requirements (publication of conformity declarations, of test reports, of audit findings, so that independent analysis can identify gaps); and structural independence of standards-setting from industry (where feasible, public standards bodies whose membership is not dominated by regulated parties).

\section{Regulatory capture}\pathtag{standard}

Regulatory capture, introduced in Chapter 10, returns here as the dynamic that explains the recurrence of regulatory-failure patterns across decades and industries. The mechanisms (expertise dependence, funding dependence, career pipeline, industry-led standard-setting, political pressure) operate in any regime in which the regulator is small relative to the regulated industry, the regulated industry is concentrated in its political and economic influence, and the technical complexity of the regulated activity makes the regulator's independent technical judgment difficult to maintain.

The structural conditions that protect against capture:

\textbf{Funding independence.} The regulator's budget is appropriated independently of the regulated industry's political activity and does not depend on the industry's fees in ways that produce dependence. Independence is partial in most arrangements; pure independence is rare. The NRC's user-fee structure (the operating licensees pay the bulk of the NRC's budget through annual fees) is an example of partial dependence whose effects are partially mitigated by the structural arrangements (the fees are set by regulation, not by industry negotiation, and the NRC's enforcement decisions are not affected by their effect on the fee base in the same year).

\textbf{Technical capability.} The regulator employs enough engineers, scientists, and analysts to perform substantive technical review independently of the industry. The capability is expensive; the budget pressure that produces capture pressure also produces capability erosion. The post-737-MAX reforms of the FAA included increases in the FAA's technical staffing as well as changes to the ODA arrangement.

\textbf{Institutional independence.} The regulator reports to a body (a board of governors, a parliament, a head of state) that is itself accountable to the public rather than to the regulated industry. The NTSB's independence from the FAA (the NTSB investigates accidents and issues recommendations; the FAA has the regulatory authority to act on them) is a structural separation that prevents the agency that certified the design from being the agency that determines whether the certification was inadequate.

\textbf{Multiple regulators.} The regulated activity is subject to multiple regulatory bodies whose jurisdictions overlap, so that capture of one regulator does not produce regulatory failure overall. The arrangement is sometimes inefficient (multiple bodies imposing inconsistent requirements) and is sometimes the only defence against single-regulator capture.

\textbf{Transparency.} The regulator's decisions, including the technical evidence and the reasoning, are public. Public visibility limits the regulator's flexibility to accommodate industry preferences in ways that the public would find indefensible. The FOIA arrangements that apply to US federal regulators (with redactions for proprietary and trade-secret information) and the equivalent European transparency requirements (under the European Ombudsman's oversight) are the structural mechanism.

\textbf{Independent technical commentary.} Academic engineering, industry analysts, public-interest groups, and the technical press conduct ongoing review of regulator decisions. The commentary creates a public record of arguments and counter-arguments that the regulator must respond to. The role of the academic safety-engineering community in commenting on certification decisions is one of the discipline's contributions to the broader regulatory function.

\section{Case: Boeing 737 MAX certification, Therac-25 regulatory aftermath, Volkswagen diesel-emissions defeat device}\pathtag{core}

\subsection*{Boeing 737 MAX certification}

The 737 MAX type certification was conducted under the FAA's Organisation Designation Authorisation (ODA) arrangement, which delegates substantial certification work to Boeing-employed engineers who serve as Authorised Representatives of the FAA \cite{acc:house-737max-2020}. The arrangement's intent was the efficient allocation of scarce regulatory engineering capacity; the consequence, identified by the House Committee report, was that the technical depth of the MCAS certification rested on engineers whose direct loyalty was to the Boeing program management chain.

The specific certification gaps identified by the House Committee and the JATR:

\textbf{The system safety assessment.} The MCAS safety assessment classified the failure modes by severity (catastrophic, hazardous, major, minor) and assigned design assurance levels (DAL A through D) accordingly. The MCAS authority was expanded during development; the safety assessment was not re-run to reflect the expansion; the design assurance level adequate for the original authority was inadequate for the expanded authority \cite{acc:house-737max-2020}.

\textbf{The single-AoA architecture.} The MCAS implementation took input from only one of the two AoA vanes per flight (alternating per power cycle); the architecture was inconsistent with prevailing certification practice for a function with MCAS's authority \cite{acc:jatr-737max-2019}. The certification process did not catch the inconsistency; the JATR's post-accident review identified the architecture as a finding.

\textbf{The pilot documentation and training.} MCAS was not documented in the operator manuals; pilots were not trained on MCAS as a discrete system; the AoA-disagree alert that would have warned of sensor disagreement was inoperative on most delivered aircraft due to a software issue Boeing identified in 2017 and did not disclose to the FAA until after the Lion Air crash \cite{acc:house-737max-2020}. The certification process did not require the disclosure or the documentation that the post-accident analysis identified as necessary.

\textbf{The "common type rating" premise.} The MAX's certification proceeded under the commercial premise that the aircraft would require no simulator training for 737NG-rated pilots, which constrained how MCAS could be documented and presented. The premise shaped the certification's depth; the certification's documentation did not flag the premise as a constraint on safety analysis.

The post-accident regulatory response included: the FAA's recertification of the 737 MAX with the MCAS redesign (dual-AoA input, cross-comparison, MCAS inhibit on disagreement, bounded authority, pilot documentation); the FAA's revision of the ODA arrangement (increased FAA oversight of delegated certification work; explicit FAA review of high-criticality systems); the 2020 Aircraft Certification, Safety, and Accountability Act (which legislated changes to FAA's certification process); and the international response (EASA, Transport Canada, ANAC of Brazil all conducted independent review and added additional requirements before approving return to service).

\subsection*{Therac-25 regulatory aftermath}

The Therac-25 incidents of 1985-1987 \cite{paper:leveson-turner1993} occurred under an FDA regulatory regime that had not anticipated software as a primary safety-critical component of a medical device. The Therac-25 was a Class II device approved under the 510(k) pathway as substantially equivalent to predecessor Therac models that had hardware interlocks. The software architecture was not independently reviewed; the dose-control software was not independently tested; the failure mode (a race condition in the software interlock fast-data-entry sequence) was not detected in pre-market evaluation.

The post-Therac-25 regulatory evolution was substantial. The Safe Medical Devices Act of 1990 expanded reporting requirements for adverse events and required manufacturers to track and investigate device problems. The FDA developed general principles of software validation, formalised in guidance documents through the 1990s and 2000s, that became the basis for software-in-medical-devices oversight. The consensus standards IEC 62304 (medical-device software life-cycle processes) and IEC 60601-1 (general electrical safety) were developed and recognised by FDA; subsequent device approvals depended on compliance with these standards. The FDA's Center for Devices and Radiological Health expanded its software-review staffing and capability.

The lesson for the regulatory regime is that the 1980s certification framework was not adequate to the technology that had emerged. The regime had been designed for an era when software was a small part of a device's safety case; by the time of Therac-25, software was the safety case for many devices, and the regime had not adapted. The regulatory failure was not the FDA's individual decision-making on Therac-25 but the structural lag of the regulatory framework behind the technology it was regulating.

\subsection*{Volkswagen diesel-emissions defeat device}

The Volkswagen diesel-emissions defeat device, in service from approximately 2008 through 2015, was engine-control software in approximately 11 million Volkswagen Group diesel vehicles worldwide that detected the test conditions of the EPA and CARB emissions test (specific patterns of throttle, steering, and dynamometer state) and altered the engine's emissions-control behaviour for the test's duration \cite{acc:epa-vw-nov-2015}. The vehicles produced certified-compliant NO$_x$ emissions on the dynamometer and non-compliant emissions (up to 40 times the limit) on the road. The device was discovered by the International Council on Clean Transportation (ICCT) in partnership with West Virginia University in 2014, through on-road emissions testing that was outside the normal regulatory protocol.

The regulatory analysis identifies several patterns. The certification regime relied on dynamometer testing under specified conditions; the regime did not include on-road testing capable of detecting test-time behaviour modification. The self-certification arrangement (Volkswagen submitted certified emissions data; the EPA accepted the certification absent contrary evidence) provided no mechanism to detect the deliberate deception. The regulator's enforcement capability after detection was substantial (the \$4.3 billion criminal-fraud settlement; the \$14.7 billion civil settlement; the executive prosecutions; the buyback program for affected vehicles) but the enforcement was reactive; the deception had operated for years before discovery.

The regulatory response included the EPA's expansion of on-road emissions testing (Portable Emissions Measurement Systems are now part of US emissions testing); the European introduction of Real Driving Emissions (RDE) testing as a complement to the laboratory cycle; the broader regulatory recognition that software-defined emissions controls require software-aware regulatory frameworks; and the ICCT's role as an independent technical authority capable of detecting regulatory evasion that the regulator's own testing could not.

The lessons aggregate: a certification regime that depends on standardised tests can be gamed by manufacturers who control the software that detects the tests; the defence against gaming is testing under conditions the manufacturer does not know in advance; independent technical capability (ICCT, academic engineering, public-interest groups) provides the surveillance the regulator's own capability cannot; and the enforcement consequences of regulatory deception must be substantial enough to deter the deception, which in turn requires the regulator's ability to detect.

\section{What the engineer owes the regulator and vice versa}\pathtag{core}

The engineer-regulator relationship is professional, not adversarial. The engineer's obligations include:

\textbf{Honest evidence.} The test reports, analysis reports, and design documentation submitted to the regulator must be accurate. Selective reporting, suppression of unfavourable results, and falsification of data are professional violations regardless of whether they are detected; they corrode the regulatory function on which public safety depends.

\textbf{Disclosure of known issues.} Defects, anomalies, and non-conformances that bear on the safety of the certified product must be disclosed to the regulator on the timescale the regulations require, regardless of the commercial consequences. The 737 MAX AoA-disagree-alert non-installation issue, which Boeing identified in 2017 and did not disclose to the FAA until after the Lion Air crash, is the negative example \cite{acc:house-737max-2020}.

\textbf{Technical contribution.} The engineer contributes to the regulatory function through participation in standards committees, in rulemaking comment periods, in industry peer-review activities, in academic and professional commentary. The contribution is part of the profession's collective work to keep the regulatory framework adequate to the technology.

\textbf{Refusal of unsafe certifications.} The engineer who is asked to certify a design they believe is unsafe is expected to refuse, to escalate the concern, and where necessary to use external channels (whistleblower protections, professional-body channels, the regulator's own confidential reporting). The codes of ethics of the engineering professional bodies (IEEE, ASME, ASCE, NSPE) all include obligations of this character.

The regulator's obligations include:

\textbf{Technical competence.} The regulator must employ enough engineers and analysts to perform substantive review and must invest in their continuing technical development. A regulator that cannot perform substantive review is a rubber stamp.

\textbf{Process integrity.} The regulator's decisions must follow the documented process; deviations from the process must be documented and justified. Process integrity is what makes regulator decisions legally defensible and publicly accountable.

\textbf{Independence.} The regulator must maintain independence from the regulated industry in funding, in personnel decisions, in technical judgment, and in enforcement decisions. Independence is not absolute (the regulator must understand the industry to regulate it), but it must be structural.

\textbf{Transparency.} The regulator's decisions, the underlying technical evidence, and the reasoning must be public to the extent that proprietary and security considerations allow. Transparency is what enables external technical review to catch regulator errors.

\textbf{Enforcement.} The regulator must have, and must use, meaningful enforcement authority. A regulator whose enforcement is symbolic is a regulator whose decisions can be ignored.

\textbf{Continuous improvement.} The regulator must learn from accidents and near-misses and must update its standards, processes, and capabilities accordingly. The lag between accident and regulatory response is a recurring failure mode; the discipline of post-accident regulatory response is the institutional version of the engineering discipline of root-cause analysis.

\begin{principle}[The regulator is not optional]
The regulatory function is part of the engineering profession's social contract. Engineering work that operates outside regulation, or that operates within regulation but treats the regulator as an obstacle to be managed rather than as a partner in the safety responsibility, undermines the social licence that allows engineering to act in domains with serious public-safety consequences. The engineer's professional self-conception includes the regulator as a legitimate counter-party.
\end{principle}

\begin{archetype}[failure]
Regulatory failure is the slow-time, structural expression of the failure archetype. The regulator's standards lag the technology; the certification regime does not catch the failure mode; the post-market surveillance does not detect the field problem; the enforcement does not deter the deviation. The accident occurs in the gap between what the regulator has established and what the technology requires.
\end{archetype}

\begin{project}[Read one accident report and critique]
Select one accident investigation report from a major national or international authority (NTSB, AAIB, BEA, JTSB, ATSB, KNKT, the IAEA, the CSB, the NRC, or analogous). Read the report end to end. Write a one-page critique that answers: (1) What was the technical mechanism the report identifies? (2) What was the organisational or regulatory mechanism the report identifies? (3) What recommendations did the report make? (4) What evidence is there in the public record that the recommendations were implemented? (5) What is missing or weak in the report from your perspective (something the report should have addressed and did not, or addressed inadequately)? Deliverable: the one-page critique, plus the report's full citation.

\textbf{Hazard.} The report is the most authoritative source on the accident; your critique is not a replacement for the report's authority but a contribution to the broader work of learning from the accident. The professional register is the same one the academic safety-engineering literature uses in commenting on investigation reports.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
The 737 MAX certification involved approximately 11,000 ODA-delegated certification activities and approximately 4,000 FAA-retained activities (representative figures from the House Committee report). Compute the ratio. Identify the categories of activity that an FAA-retained-only certification regime would either need to delegate, expand FAA staffing for, or eliminate from the certification scope.
\end{exercise}

\begin{exercise}
The Volkswagen defeat device produced certified NO$_x$ emissions of $\approx 0.04\,\text{g/mi}$ on the test cycle and on-road emissions up to $\approx 1.6\,\text{g/mi}$ \cite{acc:epa-vw-nov-2015}. Compute the ratio. Compare to the EPA limit (Tier 2 Bin 5 standard) of $0.07\,\text{g/mi}$ for the relevant period.
\end{exercise}

\begin{exercise}
The post-Therac-25 medical-device software regulatory evolution added approximately 20 years between Therac-25 (1985-1987) and the broad adoption of IEC 62304 (~2006). Compute the elapsed time. Identify what the regulatory evolution would have needed to accomplish in 5 years rather than 20 to prevent recurrent software-medical-device failures during the intervening period.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the expected number of self-certification non-conformities the regulator's audit program will detect as a function of the audit rate (fraction of declarations audited per year), the true non-conformity rate, and the detection probability per audit. Identify the audit rate at which the deterrence effect dominates.
\end{exercise}

\begin{exercise}
For a delegated-certification arrangement (ODA-style), derive a model of the conflict of interest the delegated representative faces. Identify the structural changes (oversight, separation, audit) that reduce the conflict's effect on decisions.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the cost of the FAA's full re-certification of the 737 MAX (the recertification process from grounding to return-to-service, including FAA, EASA, Transport Canada, and other national reviews). Compare to the cost of the original certification.
\end{exercise}

\begin{exercise}
Estimate the cost of independent technical capability (ICCT-style or academic emissions testing) that could detect another diesel-emissions-style defeat device. Compare to the cost of the EPA's regulatory enforcement against Volkswagen.
\end{exercise}

\begin{exercise}
Estimate the fraction of major regulatory standards (FAA airworthiness; FDA medical-device classifications; NRC reactor safety standards; EPA emissions standards) that were updated in direct response to specific named accidents. Discuss what the fraction implies about regulator-driven versus accident-driven regulatory evolution.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a regulatory regime for autonomous-vehicle software safety. Specify the standard-setting approach, the pre-market evaluation approach, and the post-market surveillance approach. Identify the analogues from the existing regimes (aviation, medical devices, nuclear, automotive) and where the autonomous-vehicle case requires something new.
\end{exercise}

\begin{exercise}
Design the post-737-MAX reform of FAA's ODA arrangement. Specify what oversight FAA must retain, what can be delegated, what audit and review mechanisms apply, and how the delegated representatives' independence is protected. Justify each design choice against the House Committee findings.
\end{exercise}

\begin{exercise}
Design an enforcement framework for a self-certification regime that produces the deterrence-effect strength of a pre-market-evaluation regime. Specify the audit rates, the penalties, the disclosure requirements, and the criminal-referral thresholds.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: the ODA arrangement is structurally incapable of producing safe certifications for systems with the complexity of modern airliners, and the post-737-MAX reforms do not change the structural incapability. Argue the opposite position. Identify the empirical evidence that bears on the dispute.
\end{exercise}

\begin{exercise}
Discuss the Therac-25 regulatory aftermath \cite{paper:leveson-turner1993}. Identify what the FDA could have done in the 1980s, with the resources and authority it had, that would have caught the software architecture's hazards before the device entered service.
\end{exercise}

\begin{exercise}
Argue: the Volkswagen defeat device is evidence that the engineering profession's self-regulation through codes of ethics is insufficient, and stronger external regulation is required \cite{acc:epa-vw-nov-2015}. Identify the changes to engineering education, professional registration, or institutional accountability that would address the gap.
\end{exercise}

\begin{exercise}
Discuss the gap between the regulator's needed technical capability and the regulator's funded technical capability for any of the major regulatory regimes (FAA, FDA, NRC, EPA). Identify the structural causes of the gap and the political constraints on closing it.
\end{exercise}
