\chapter{Corrosion and degradation}
\label{vol05:ch09}

\epigraph{Corrosion is the destruction or deterioration of a material
because of reaction with its environment. The reaction is
electrochemical for metals in solution, chemical for polymers in
ultraviolet light, mechanical for any material under cyclic wear. The
engineering of materials in service is the engineering of these
reactions against the design service life.}{Adapted from Mars G.\
Fontana, working scope of corrosion engineering \cite[ch.~1, pp.~1--4]{
text:fontana-corrosion}.}

\chapmeta{Half-life: foundational. Archetypes: failure, transport. Exercise target: 32. Project track: physical measurement (corrosion coupons). Hazard class: medium (acidic and saline corrosion solutions; ventilated handling). Reader path default: standard, with sections explicitly tagged. Named cases: Aloha Airlines Flight 243 (1988); Hyatt Regency walkway connection (1981) referenced for galvanic-couple context; chloride-induced reinforcement corrosion in concrete bridge decks described by mechanism only; ageing-aircraft and marine-structure cases described by mechanism without naming non-registered events.}

\noindent
A material in service is a material under attack. The attack is
electrochemical for metals (corrosion by aqueous, atmospheric, and
high-temperature mechanisms), mechanical for contact surfaces
(wear in three abrasive, adhesive, and erosive modes), nuclear for
materials in radiation fields, and ultraviolet-and-thermal for
polymers. The chapter develops each mechanism on engineering footing:
the rate law, the controlling variables, the diagnostic, the
protective intervention. The closing section names three working
failure modes (corrosion-fatigue in aircraft skin, chloride-induced
spalling in reinforced concrete, stress-corrosion cracking in
chemical-plant piping) that the engineer must screen for in any
materials-selection review. The chapter is the discipline-side
counterpart to Volume V Chapter 10: selection chooses the material;
this chapter is honest about what the environment will do to it.

\section{Corrosion mechanisms: galvanic, pitting, crevice, intergranular}\pathtag{core}

A metal in an aqueous environment is at electrochemical equilibrium
with the solution at a potential set by its tendency to give up
electrons. A metal that gives up electrons more readily than another
(more negative standard potential) corrodes preferentially when the
two are coupled in a conducting electrolyte. The full electrochemical
machinery of Volume V Chapter 3 (the Nernst equation, the standard
electrode-potential series) carries over directly; this chapter
focuses on the engineering forms.

\engterm{The eight forms of corrosion}. Fontana's canonical taxonomy
\cite[ch.~3]{text:fontana-corrosion} identifies eight distinct
engineering forms, each with its own mechanism, diagnostic, and
remediation:

\begin{itemize}[leftmargin=*]
\item \engterm{Uniform attack}: the metal corrodes evenly over the
  exposed surface. The corrosion rate, expressed as a thickness loss
  per year ($\text{mm}/\text{y}$ or, in U.S. practice, mils per year,
  $\text{mpy}$), is the working figure. Mild steel in atmospheric
  exposure: $0.05$--$0.5\,\text{mm}/\text{y}$ depending on humidity
  and chloride content. Stainless steel in clean atmosphere: below
  $0.001\,\text{mm}/\text{y}$.
\item \engterm{Galvanic corrosion}: two dissimilar metals in
  electrical contact in an electrolyte. The less noble metal (more
  negative potential) corrodes preferentially. The galvanic series
  in seawater orders practical alloys: from anodic (magnesium, zinc,
  aluminium alloys, mild steel, cast iron) to cathodic (brass,
  copper, stainless steel passivated, titanium, graphite, gold).
\item \engterm{Crevice corrosion}: localised corrosion in narrow gaps
  between two metal surfaces or between a metal and a non-metallic
  cover (gasket, deposit, biofilm). The crevice limits diffusion;
  oxygen is depleted inside the crevice; the metal inside becomes
  anodic to the metal outside. Stainless steels are particularly
  susceptible, because crevice-restricted oxygen breaks down the
  protective passive film.
\item \engterm{Pitting corrosion}: localised attack that produces
  deep, narrow holes in an otherwise sound surface. Pitting initiates
  at sites of compositional or microstructural inhomogeneity (an
  inclusion, a grain-boundary precipitate); the pit grows by
  autocatalytic acidification (the chloride and metal ion content
  inside the pit drive the pit deeper). The working threat is that
  the pit may penetrate the wall thickness with little loss of mass;
  inspection by mass loss alone will miss the failure.
\item \engterm{Intergranular corrosion}: preferential attack along
  grain boundaries. The classical case is sensitised austenitic
  stainless steel: at $500$--$800\,\degC$, chromium carbides
  precipitate at grain boundaries and deplete the adjacent matrix of
  chromium below the $\sim 11\,\%$ minimum for passivity. The depleted
  zone corrodes; the grains fall out as discrete entities. The
  remediation is to use low-carbon ($< 0.03\,\%$) or stabilised
  (titanium-, niobium-modified) stainless grades that suppress
  carbide precipitation.
\item \engterm{Selective leaching} (dealloying): one component of an
  alloy preferentially corrodes. The canonical case is dezincification
  of brass: zinc dissolves preferentially, leaving a porous, brittle
  copper-rich residue. Graphitic corrosion of cast iron is the
  analogous process for iron: iron dissolves, leaving the graphite
  flakes behind. The remaining material has the shape of the original
  part but the strength of porous copper or graphite.
\item \engterm{Erosion-corrosion}: combined corrosion and mechanical
  erosion. The fluid flow removes the protective corrosion product
  film as fast as it forms; the metal corrodes at a higher rate than
  in still solution. Working examples: pump impellers, valve seats,
  elbow bends in pipework, condenser tubes downstream of partial
  blockages. The remediation is to keep the flow regime below the
  threshold velocity for the chosen material.
\item \engterm{Stress-corrosion cracking}: the eighth form, treated in
  section~9.2.
\end{itemize}

\engterm{Pourbaix diagrams}. The thermodynamic stability of a metal
in solution as a function of potential and pH is given by the Pourbaix
diagram, a two-dimensional plot of regions of immunity (metal is
thermodynamically stable as metal), corrosion (metal is stable as
soluble cation), and passivation (metal is stable as a protective
oxide film). The diagram shows where the metal can be protected by
adjusting potential (cathodic protection moves the system into
immunity) or by adjusting pH (alkaline solutions move iron into the
passive region). Pourbaix diagrams are computed from thermodynamic
data; they are necessary but not sufficient: a metal in a passive
region may still corrode by mechanisms (pitting, crevice) that the
thermodynamic diagram does not predict \cite[ch.~2]{text:jones-corrosion}.

\engterm{Working corrosion rates}. The corrosion current density
$i_{\text{corr}}$ converts directly to thickness loss by Faraday's
law. For iron ($\rho = 7870\,\text{kg}/\text{m}^3$, equivalent weight
$M/z = 27.9\,\text{g}/\text{eq}$),
\[
\text{rate} = \frac{i_{\text{corr}} M}{z F \rho} = i_{\text{corr}} \times
1.16 \times 10^{-2}\,\text{mm/year per }\si{\micro\ampere\per\centi\meter\squared}.
\]
A corrosion current density of $10\,\si{\micro\ampere\per\centi\meter\squared}$
corresponds to $0.12\,\text{mm}/\text{y}$ of iron loss. The working
range for corrosion currents is $10^{-3}$ to $10^3\,\si{\micro\ampere\per\centi\meter\squared}$,
covering all engineering corrosion situations from passive stainless
steel to actively dissolving zinc anode.

\begin{archetype}[Failure and transport]
A corrosion failure is the failure mode whose rate is set by the
transport of reactants to the metal surface (oxygen, chloride, water,
or the electron-rejecting cathodic reaction) and by the chemistry that
turns those reactants into the corrosion product. The engineering
analysis is two-step: identify the binding transport process (oxygen
diffusion through paint, chloride diffusion through concrete cover,
electrolyte film thickness in atmospheric corrosion, fluid renewal
rate in flowing systems), and identify the chemistry at the metal
surface that the transport supplies. Every corrosion remediation
intervenes on one or the other: paint slows oxygen diffusion, cathodic
protection drives the chemistry into immunity, alloy choice changes
the surface response to the supplied reactant.
\end{archetype}

\section{Stress-corrosion cracking}\pathtag{standard}

A metal under tensile stress in a specific corrosive environment can
fail by cracking at stresses well below its yield strength, on
timescales of days to years. The combination of stress and environment
is the necessary condition; either alone is insufficient. The working
examples are notorious:

\begin{itemize}[leftmargin=*]
\item Austenitic stainless steel in hot chloride solutions: the most-
  cited case. Type 304 and 316 stainless steel above $\sim 60\,\degC$
  in chloride-containing water (industrial cooling water, swimming-
  pool atmospheres, marine atmospheres) crack transgranularly under
  modest tensile stress. The remediation: higher-nickel grades (Alloy
  825, Alloy 20) or duplex stainless (2205, 2507), or lower
  operating temperatures, or removal of the chloride.
\item Brass in ammonia atmospheres: ``season cracking'' of cold-
  worked brass cartridge cases, documented since the British colonial
  garrisons in India in the 19th century. Cracking initiates at
  residual tensile stresses from cold work; remediated by stress-
  relief annealing or by alloy substitution.
\item Carbon steel in caustic solutions: caustic embrittlement of
  steam boilers, the failure mode that drove much of the early
  20th-century boiler-explosion record. Remediated by deaeration of
  feedwater, control of caustic concentration, and by stress-relief
  of welded joints.
\item High-strength aluminium alloys (7075-T6, 7079) in chloride
  atmospheres: limits the use of the highest-strength aerospace alloys
  in primary structure that contacts the environment.
\item High-strength steels in hydrogen-bearing environments (sour gas,
  cathodic protection of high-strength fasteners): hydrogen-assisted
  cracking is mechanistically related; the remediation is to limit
  yield strength below $\sim 1100\,\text{MPa}$ and to bake out
  absorbed hydrogen after electroplating.
\end{itemize}

\engterm{The threshold stress intensity for SCC}. The stress-intensity
factor for a crack in a specific environment must exceed a threshold
$K_{\text{ISCC}}$ for SCC propagation to occur. Below the threshold,
the crack does not advance (within the observation window); above it,
the crack advances at a rate that depends on $K_I$ and on the
environment. The working design check is that the maximum service
$K_I$, including residual stresses, must lie below $K_{\text{ISCC}}$
for the chosen material and environment. Working values:
$K_{\text{ISCC}}/K_{Ic} \sim 0.1$--$0.3$ for the susceptible alloys
in their susceptible environments; the SCC threshold is well below
the static fracture toughness.

\engterm{The three-leg diagnostic}. An SCC investigation establishes
three legs: a susceptible material, a corrosive environment, and a
sustained tensile stress (applied or residual). Removing any leg
prevents the failure. The remediation chosen in practice is whichever
leg is cheapest to remove: stress-relief heat treatment (remove
residual stress); inhibitor addition or environment change (remove
the corrodent); alloy substitution (remove the susceptibility).

\engterm{Corrosion fatigue}. The combination of corrosion and cyclic
loading produces accelerated crack growth: the corrosive environment
weakens the metal at the crack tip and the cyclic load advances the
crack faster than either alone. The S-N curve for a corroded material
loses its endurance limit (the asymptotic flat region at $\sim 10^7$
cycles in dry-air S-N curves); fatigue strength continues to decrease
with cycle count. The Paris law for crack-growth rate
$da/dN = C(\Delta K)^m$ has a larger pre-exponential $C$ in corrosive
environments, often by a factor of $5$--$20$. The Aloha Airlines 243
case (section~9.7) is the canonical corrosion-fatigue failure.

\section{Protection: coatings, cathodic protection, alloy choice, design}\pathtag{standard}

The engineering management of corrosion uses four families of
intervention.

\engterm{Alloy choice}. The single most effective intervention is to
choose a material that does not corrode in the service environment.
Stainless steels (chromium $> 11\,\%$ for passive oxide), nickel-based
alloys (Inconel, Hastelloy), titanium (passive oxide stable across
oxidising environments), and aluminium (passive oxide on initial
exposure, then protective). The cost of more corrosion-resistant
alloys is typically two to ten times the cost of mild steel for the
same strength; the working balance is between corrosion-allowance
thickness on mild steel and substitution to a higher-grade alloy. The
balance shifts in favour of substitution as the design service life
exceeds $\sim 20$ years, the inspection burden grows, or the
consequence of failure is high.

\engterm{Coatings}. The coating separates the metal from the
environment. Working coating families:

\begin{itemize}[leftmargin=*]
\item \engterm{Organic coatings} (paints, varnishes, two-pack epoxy
  systems): physical barriers to oxygen and water. Working lives
  range from a few years (consumer paint) to over forty years
  (industrial epoxy coatings with high-build cycles). Failure modes
  are pinholes, mechanical damage, and adhesion loss at the substrate
  interface.
\item \engterm{Metallic coatings}: a more corrosion-resistant metal
  applied as a layer on the substrate. Zinc on steel (galvanising)
  is the canonical case: zinc is anodic to steel in most environments,
  so a scratch through the coating does not lead to undercutting (the
  steel is cathodically protected). Aluminium-zinc coatings (Galfan,
  Galvalume) extend the life. Cadmium plating on aerospace fasteners
  combines barrier and sacrificial protection.
\item \engterm{Conversion coatings}: chemical reaction with the
  substrate to form a thin protective layer. Phosphating before
  painting (improves paint adhesion), anodising of aluminium (thickens
  the natural oxide film), chromating (now restricted under REACH and
  RoHS regulations because of hexavalent-chromium toxicity).
\item \engterm{Inorganic coatings}: porcelain enamel, glass linings,
  refractory cement. Long working lives in chemically aggressive
  environments at modest cost; brittle, so vulnerable to thermal and
  mechanical shock.
\end{itemize}

\engterm{Cathodic protection}. The metal is forced to be a cathode by
external means. Two methods:

\begin{itemize}[leftmargin=*]
\item \engterm{Sacrificial anode}: a more active metal (zinc,
  aluminium, magnesium) is electrically connected to the protected
  structure. The anode corrodes preferentially; the structure is
  protected. Zinc anodes on ship hulls and on heating-system steel
  tanks; magnesium anodes on buried steel pipelines in low-resistivity
  soils. Sacrificial-anode systems are passive (no external power);
  the anode is consumed at a rate proportional to the protective
  current and must be replaced.
\item \engterm{Impressed current}: an external DC power supply drives
  electrons to the protected structure from an inert auxiliary anode
  (high-silicon cast iron, mixed-metal-oxide-coated titanium, lead-
  silver alloys). Impressed-current systems protect large structures
  (oil-and-gas pipelines, harbour structures, large tanks); the
  capital cost is higher than sacrificial-anode systems, the operating
  cost includes the electrical energy, and the protective current
  can be tuned to the structure's needs.
\end{itemize}

The working protection criterion is to depress the metal's potential
to $-0.85\,\text{V}$ (versus Cu/CuSO$_4$ reference electrode) for
iron in soil, into the immunity region of the Pourbaix diagram. The
NACE SP0169 standard (Recommended Practice for Control of External
Corrosion on Underground or Submerged Metallic Piping Systems) sets
the working criterion.

\engterm{Design for corrosion}. The fourth intervention is the design
itself. Working principles:

\begin{itemize}[leftmargin=*]
\item Avoid crevices, lap joints, and gasket pockets that retain
  water. Tank bottoms welded flat to the floor rather than lapped;
  drainage holes at low points; sealant beads at lap joints.
\item Avoid galvanic couples between dissimilar metals; if a couple
  is unavoidable, electrically isolate the components with an
  insulating washer or a non-conducting gasket.
\item Avoid sharp internal corners (stress concentration plus crevice
  for deposit). Generous fillet radii; smooth-flow internal surfaces.
\item Place dissimilar-metal fasteners and structural connections in
  locations that are easy to inspect and replace. The Hyatt Regency
  walkway connection \cite{acc:nbs-hyatt-1982} was a single load-
  bearing fastener whose modification was not inspected; the
  corrosion-management principle is the same: a structural element
  whose replacement on the available inspection cycle is impractical
  should be over-designed against the binding degradation mode.
\item Specify the corrosion allowance explicitly in the design (the
  additional thickness over the strength-required thickness, intended
  to be consumed by uniform corrosion over the design life). Working
  values: $1.5\,\text{mm}$ for refinery vessels in low-corrosion
  service, $3\,\text{mm}$ for moderate-corrosion service, $6\,\text{mm}$
  or more for high-corrosion service.
\end{itemize}

\section{Wear: abrasive, adhesive, erosive}\pathtag{standard}

A material in mechanical contact with another material loses substance
through wear. The three primary mechanisms are abrasive, adhesive, and
erosive wear; secondary mechanisms (fatigue wear, corrosive wear,
fretting) combine these with cyclic loading or chemical attack.

\engterm{Abrasive wear}. Hard particles or asperities cut grooves in
the softer surface. Two-body abrasion: the abradant is fixed to the
counterface (a grinding wheel against a workpiece); three-body
abrasion: the abradant moves freely between the surfaces (sand grit
between brake disc and pad). The Archard wear law gives the working
relation
\[
V = K \frac{F L}{H},
\]
with $V$ the volume worn, $F$ the normal load, $L$ the sliding
distance, $H$ the hardness of the softer surface, and $K$ a wear
coefficient that depends on the materials and the lubrication
condition. Working $K$ values: $\sim 10^{-2}$ for unlubricated
metal-on-metal (severe wear), $\sim 10^{-4}$ for lubricated similar
metals, $\sim 10^{-6}$ for hardened steel pin-on-disc in light
load, $\sim 10^{-8}$ for polymer-on-metal at low load.

\engterm{Adhesive wear}. Microscopic junctions form at points where
asperities contact under load; the junctions weld locally and shear
under relative motion, transferring material from one surface to the
other. Galling, scoring, and seizure are extreme cases. Adhesive wear
is suppressed by hard, chemically dissimilar surfaces; soft coatings
(soft-metal coatings on hard substrates, polymer coatings on hard
substrates); and lubrication that separates the contacting surfaces.

\engterm{Erosive wear}. Solid particles or liquid droplets entrained
in a fluid stream impinge on a surface and remove material by
impact. The wear rate depends on the impact angle: brittle materials
peak at $90^\circ$ (normal impact, crater formation); ductile
materials peak at $\sim 20$--$30^\circ$ (cutting-action removal of
material). Working examples: turbocharger blades in dust-laden intake
air, hydraulic-turbine runner buckets in silty water, pneumatic-
conveyor elbows in particulate-laden gas, rocket-engine nozzles in
particulate-laden exhaust.

\engterm{Wear-resistant materials and coatings}. The hardness $H$ in
Archard's law is the lever. Engineering approaches:

\begin{itemize}[leftmargin=*]
\item Bulk hard materials: hardened steels, cast irons, tool steels.
\item Hard coatings: tungsten carbide (WC), titanium nitride (TiN),
  chromium carbide (Cr$_3$C$_2$), diamond-like carbon (DLC), applied
  by thermal spray, CVD, PVD, or weld overlay.
\item Composite surfaces: WC particles in a ductile cobalt matrix
  (the cutting-tool inserts); ceramic particles in a metal-matrix
  composite coating (laser-clad MMCs on excavator buckets).
\item Surface hardening of bulk metals: case carburising, nitriding,
  induction hardening, laser hardening. Section~11 of this volume
  develops the manufacturing-side metallurgy.
\end{itemize}

\engterm{Lubrication}. Lubricants separate the contacting surfaces.
Three regimes are distinguished by the ratio of lubricant film
thickness $h$ to combined surface roughness $\sigma$:

\begin{itemize}[leftmargin=*]
\item \engterm{Boundary lubrication} ($h/\sigma < 1$): direct asperity
  contact through the lubricant. Working friction coefficients
  $\sim 0.1$.
\item \engterm{Mixed lubrication} ($1 < h/\sigma < 3$): partial
  asperity contact; friction coefficient $\sim 0.05$--$0.1$.
\item \engterm{Hydrodynamic} or \engterm{elastohydrodynamic
  lubrication} ($h/\sigma > 3$): no asperity contact; friction
  coefficient $\sim 0.001$--$0.01$.
\end{itemize}

The working journal bearings of automobile engines operate in the
hydrodynamic regime; the rolling-element bearings operate in the
elastohydrodynamic regime; the brake pads operate in boundary
contact.

\section{Radiation damage}\pathtag{mastery}

Materials in radiation fields (the cores of fission reactors, the
first walls of fusion reactors, the structural components of
spacecraft, the storage casks of spent fuel, the medical-imaging
shielding) develop changes in microstructure and mechanical properties
that drive their service life.

\engterm{Mechanisms}. Energetic particles (neutrons, ions, electrons,
photons) collide with atoms in the lattice and displace them from
their crystal sites. The primary knock-on atom may carry enough
energy to displace further atoms; a cascade of $\sim 10^3$ -- $10^5$
displacements forms from a single fast neutron, in a region of a few
nanometres on a timescale of a few picoseconds. The cascade leaves
behind a population of point defects: vacancies, interstitials,
and (in many alloys) gas atoms (helium from $(n,\alpha)$ reactions
in nickel-bearing steels, hydrogen from $(n,p)$).

\engterm{Working units}. Radiation damage is quantified by
\engterm{displacements per atom} (dpa): the average number of times
each atom in the material has been displaced from its lattice site
during exposure. A pressurised-water-reactor pressure vessel
accumulates $\sim 0.01$ -- $0.1\,\text{dpa}$ over its $60$-year
design life; a fast-reactor cladding $\sim 100\,\text{dpa}$; a
proposed fusion-reactor first wall $\sim 100\,\text{dpa}$ over a
five-year campaign.

\engterm{Mechanical-property changes}. The defect population produces
several engineering effects:

\begin{itemize}[leftmargin=*]
\item \engterm{Hardening and embrittlement}: defect clusters and
  precipitates pin dislocations; yield strength rises, ductility falls.
  In ferritic-steel reactor pressure vessels, the ductile-to-brittle
  transition temperature shifts upward with dpa; the shift is the
  working measure of vessel life.
\item \engterm{Swelling}: voids nucleate and grow from accumulated
  vacancies; the material expands at constant mass. Austenitic
  stainless cladding in fast-reactor service can swell $\sim 10\,\%$
  by volume; the swelling distorts fuel-pin geometry and is the
  binding life-limiting mechanism for the cladding.
\item \engterm{Creep enhancement}: irradiation provides excess
  vacancies; diffusional creep mechanisms (Nabarro-Herring, Coble)
  are accelerated. Irradiation creep is the working life-limiter for
  zirconium-alloy fuel cladding in light-water reactors.
\item \engterm{Phase changes and segregation}: radiation drives
  non-equilibrium phase transformations; solutes segregate to grain
  boundaries (radiation-induced segregation); precipitates dissolve
  or coarsen.
\end{itemize}

\engterm{Polymer behaviour in radiation}. Polymers degrade by chain
scission (random breaking of the polymer backbone, reducing molecular
weight and tensile strength) or crosslinking (formation of bonds
between chains, increasing rigidity and reducing ductility). The
dominant mechanism depends on the polymer family: polyethylene and
polyimides crosslink under radiation; polypropylene, PTFE, and
polysulphone scission. Working dose limits for engineering polymers
in service: $\sim 10^4$ -- $10^6\,\text{Gy}$ for ductility-critical
applications, three to four orders of magnitude below the working
dose limits for engineering steels.

\engterm{Engineering responses}. The materials choice for high-
radiation environments is dominated by a small set of alloys
specifically qualified by long-duration test irradiation. For
pressurised-water reactors: low-copper, low-nickel ferritic steels
(SA508 forgings and SA533B plates) and zirconium alloys for cladding.
For fast reactors: ferritic-martensitic steels (HT9, T91, T92) for
ductless designs and oxide-dispersion-strengthened steels for
high-temperature claddings. For proposed fusion reactors: reduced-
activation ferritic-martensitic steels (Eurofer-97, F82H) and silicon-
carbide-matrix composites for the first wall.

\section{Polymer ageing revisited}\pathtag{standard}

Polymers age by several mechanisms documented in Volume V Chapter 6
(polymers) and revisited here in the context of corrosion-and-
degradation engineering.

\engterm{Photo-oxidation}. Ultraviolet light, particularly UV-A
($315$--$400\,\text{nm}$) and UV-B ($280$--$315\,\text{nm}$) at the
earth's surface, has photon energies of $3.1$--$4.4\,\text{eV}$,
comparable to bond-dissociation energies. UV absorption in the polymer
backbone initiates a free-radical chain that consumes oxygen and
breaks the chain. The visible effects are chalking (loss of surface
gloss), yellowing or colour change, embrittlement, and crack
formation. The remediation: UV stabilisers (hindered amine light
stabilisers, HALS; UV absorbers; pigments such as carbon black or
titanium dioxide).

\engterm{Thermal ageing}. Polymers at elevated temperature undergo
thermo-oxidative chain scission and crosslinking at rates following
Arrhenius behaviour, $k \propto \exp(-E_a/RT)$. The working figure is
the temperature-life relation: the time to reach a $50\,\%$ loss of
some property (tensile strength, elongation at break) doubles for
every $10\,\degC$ decrease in service temperature, approximately. The
Arrhenius exponent is calibrated by accelerated ageing tests at
$80$--$120\,\degC$, extrapolated to service temperatures.

\engterm{Hydrolytic degradation}. Polyesters, polyamides, polyimides,
polycarbonates, and any polymer with hydrolysable ester, amide, or
carbonate links lose molecular weight by reaction with water,
particularly at elevated temperature. The mechanism is the working
life-limiter for medical-implant polymers (PLA, PLGA), for fishing-net
and rope nylon in marine service, and for aerospace polyimide laminates
in humid service.

\engterm{Environmental stress cracking}. Polymers under tensile stress
in the presence of certain chemicals (alcohols, surfactants, vegetable
oils) crack at stresses well below the yield strength of the bulk
polymer. The mechanism is analogous to stress-corrosion cracking in
metals, with the chemical agent reducing surface energy and the stress
driving crack propagation. The engineering response is to specify
polymers tested against the service-environment chemicals (the ESCR,
environmental-stress-crack resistance, tests in ASTM and ISO
standards).

\engterm{Working service lives}. Polymer service life is set by the
binding degradation mechanism and by the published dose-life curves.
Working ranges in service:

\begin{itemize}[leftmargin=*]
\item Polyethylene-pipe in buried service: $50$--$100\,\text{y}$.
\item PVC window frames in outdoor exposure: $30$--$50\,\text{y}$
  with stabiliser package.
\item EPDM roofing membrane: $20$--$30\,\text{y}$.
\item Aerospace structural composites in service: $30$--$40\,\text{y}$
  with periodic inspection.
\item Polyester-fibre rope in UV exposure: $1$--$5\,\text{y}$.
\end{itemize}

Volume V Chapter 6 develops the polymer family chemistry; this
section establishes the engineering measure of life under
environmental attack.

\section{Failure: corrosion-driven catastrophic failures}\pathtag{core}

Three documented failure patterns illustrate the working content of
corrosion-and-degradation engineering at the catastrophic end of the
consequence spectrum.

\engterm{Corrosion-fatigue in pressurised aircraft skin: Aloha
Airlines Flight 243}. On 28 April 1988, a Boeing 737-200 lost a
$5.5\,\text{m}$ section of upper-forward fuselage skin at $24{,}000\,
\text{ft}$ over Hawaii \cite{acc:ntsb-aar94-04}. The NTSB
investigation identified the failure mode as multiple-site fatigue
cracking at cold-bonded lap joints, accelerated by salt-spray
corrosion. The cold-bonded lap joints, intended to share load between
the rivet rows and the adhesive bond, had disbonded in service; the
load fell entirely on the rivet rows; salt-spray corrosion of the
aluminium at the rivet holes acted as a stress-concentration site;
fatigue cracks initiated at adjacent rivet holes and grew until they
linked up. The aircraft had accumulated $89{,}680$ flight cycles
against an original design service objective of $75{,}000$ cycles.

The canonical lessons:

\begin{itemize}[leftmargin=*]
\item The corrosion mechanism (salt-spray attack of the disbonded lap
  joint) and the fatigue mechanism (cyclic pressurisation) reinforced
  each other; neither alone would have produced the failure on this
  timescale.
\item The visual-inspection regime did not match the failure mode
  (multiple-site damage with many short cracks rather than one long
  crack). The post-Aloha FAA Aging Aircraft Program revised the
  inspection regime for cold-bonded lap joints to include eddy-current
  and ultrasonic techniques.
\item The original design intent (the cold bond carries load in
  parallel with the rivets) was not durable in service; the as-
  manufactured cold bond degraded by mechanisms (moisture, salt) that
  the design analysis did not account for. Volume V Chapter 11 returns
  to the gap between as-designed and as-manufactured performance.
\end{itemize}

\engterm{Chloride-induced corrosion of reinforcing steel in concrete
bridge decks}. The failure pattern is not a single named event but a
canonical degradation mode of essentially every reinforced-concrete
bridge deck in cold-weather climates where de-icing salt is applied.
The mechanism: chloride ions from de-icing salt or from marine spray
diffuse through the concrete cover to the reinforcing steel; chloride
depassivates the steel by breaking down the protective alkaline-oxide
film; the depassivated steel corrodes; the corrosion product is
two to six times the volume of the consumed steel; the expansive
pressure cracks and spalls the cover concrete; once the cover is
breached, oxygen and water reach the steel directly and the corrosion
accelerates. The visible signs (in order of progression): rust
staining at the bottom of the deck slab, longitudinal cracks along
the line of reinforcement, spalling of cover concrete, exposure of
reinforcement, loss of reinforcement cross-section, loss of structural
capacity.

The remediation, on order of escalation: improved concrete (low-
permeability mix, higher cement content, supplementary cementitious
materials), increased cover (the ACI 318 cover requirement is the
canonical regulatory expression), corrosion-resistant reinforcement
(epoxy-coated rebar, galvanised rebar, stainless rebar, fibre-reinforced-
polymer rebar), and post-construction cathodic protection of the
reinforcement in service. The chloride-diffusion engineering produces
working life predictions:
\[
t_{\text{init}} \approx \frac{c^2}{4 D_{\text{Cl}} \,
[\text{erfc}^{-1}(C_{\text{th}}/C_s)]^2},
\]
with $c$ the cover thickness, $D_{\text{Cl}}$ the effective chloride-
diffusion coefficient in the concrete (typically $10^{-12}$ -- $10^{-11}\,
\text{m}^2/\text{s}$), $C_{\text{th}}$ the threshold chloride
concentration at the steel for depassivation, and $C_s$ the surface
chloride concentration. For $c = 50\,\text{mm}$, $D = 10^{-12}\,
\text{m}^2/\text{s}$, threshold and surface concentrations of order
unity, the initiation time is decades; for $c = 25\,\text{mm}$ at the
same $D$, the initiation time is years.

\engterm{Stress-corrosion cracking in chemical-plant piping}. Type
304 and 316 stainless-steel piping in hot chloride-containing service
(cooling water, salt-laden steam, sea-water condensate) has cracked
in service across the chemical-process industries since the alloy's
introduction. The cracking is transgranular, branched, follows a
roughly perpendicular path to the applied tensile stress, and
proceeds at rates of millimetres per year once initiated. The
remediation on existing plants is a substitution programme:
duplex stainless (2205, 2507), nickel-base alloys (Alloy 800, 825),
or, in the worst cases, titanium. The cost of the substitution drives
the engineering decision; the cost of an in-service cracking failure
(unplanned shutdown, environmental release, personal injury) is
usually the operative comparison.

The three patterns share a common structure: a corrosion mechanism
that the original design analysis did not adequately account for, an
inspection regime that did not match the failure mode, and a remediation
that requires either material substitution or design change. Volume X
returns to several of these cases in the systems-engineering context.

\begin{principle}[Corrosion failures are predictable from the
environment-material-stress triad]
A material in service that fails by corrosion or stress-corrosion
cracking fails at the intersection of three preconditions: the
material is susceptible (composition, microstructure, residual stress),
the environment is corrosive (chemistry, temperature, transport), and
the stress is sustained (applied, residual, or cyclic). Every
corrosion-failure investigation reconstructs the triad. Every
corrosion-remediation removes one of the three legs. The discipline of
materials-in-service is the discipline of naming the triad before the
part enters service, not after.
\end{principle}

\begin{estimation}
\textbf{A buried steel water-distribution pipe is to be cathodically
protected by sacrificial magnesium anodes. The pipe has external
surface area $A = 1{,}000\,\text{m}^2$, runs through soil of
resistivity $\rho_s = 50\,\si{\ohm}\cdot\text{m}$, and requires a
protective current density $i_p = 10\,\text{mA}/\text{m}^2$. Estimate
the number of $7\,\text{kg}$ magnesium anodes required and their
service life.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The total protective current is $I = i_p A = 10 \times 1{,}000 =
10\,\text{A}$.

Each magnesium anode delivers a current that depends on the
soil-resistivity-determined circuit resistance and on the
driving potential. The driving potential between the magnesium anode
($-1.55\,\text{V}$ vs Cu/CuSO$_4$) and the protected steel
($-0.85\,\text{V}$ at full protection) is approximately $0.7\,\text{V}$.
The resistance of a single magnesium anode of typical dimensions
($\sim 0.1\,\text{m}$ diameter, $0.3\,\text{m}$ long) in soil follows
the Dwight formula
$R \approx (\rho_s/2\pi L) [\ln(4L/d) - 1]$, which evaluates to
$R \approx (50 / 6.28 \times 0.3) [\ln(12) - 1] = 26.5 \times 1.48
\approx 40\,\si{\ohm}$ per anode.

The current per anode is $I_{\text{anode}} \approx 0.7 / 40 =
0.0175\,\text{A} = 17.5\,\text{mA}$.

The number of anodes required: $N = I / I_{\text{anode}} = 10 /
0.0175 \approx 570$ anodes.

The mass of magnesium consumed per anode per year (Faraday's law,
$8.76\,\text{kg}$ of magnesium per amp-year for the divalent
$\text{Mg} \to \text{Mg}^{2+}$ reaction):
$\dot{m}_{\text{anode}} = 0.0175 \times 8.76 = 0.153\,\text{kg/y}$.

The lifetime of a $7\,\text{kg}$ anode is approximately $7/0.153
\approx 45\,\text{years}$, well over the intended pipe service life.
The binding constraint is the number of anodes, not the lifetime of
each anode.

A real installation would use fewer, larger anodes ($14$ -- $30\,\text{kg}$
each) and would group them at locations of high current demand
(coating defects, river crossings). The pipe would also be coated;
the protective current would then run an order of magnitude lower than
our $10\,\text{mA}/\text{m}^2$ figure (typical $\sim 0.5\,\text{mA}/
\text{m}^2$ for a well-coated pipe), and the anode count would
correspondingly drop.

The postmortem: the back-of-envelope captures the scaling of anode
count with surface area and soil resistivity. The actual design relies
on the coating to reduce the bare-steel demand by a factor of ten
or more, after which the anode count is manageable. The discipline
question is whether the coating's durability over the design life
(say $50$ years) is the binding life-limiter rather than the anode
mass.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Corrosion-rate measurement on three metals in three
environments]
The reader exposes three metal coupons to three different aqueous
environments, measures the mass loss over a chosen exposure interval,
and infers corrosion rates and the dominant mechanism in each case.
The deliverable is a fifteen- to twenty-five-page report.

\textbf{Track}. Physical measurement.
\textbf{Hazard class}: medium. Acidic and saline solutions; gloves,
eye protection, ventilation. Disposal of test solutions per local
waste regulations.

\textbf{Materials}.

\begin{itemize}[leftmargin=*]
\item Three metal coupons of similar surface area
  ($\sim 25 \times 50\,\text{mm}$, $\sim 1\,\text{mm}$ thick):
  mild steel, copper, aluminium. Or any three of: zinc, brass, type
  304 stainless, lead.
\item Three test solutions in $\sim 250\,\text{mL}$ jars or beakers:
  $3\,\%$ NaCl (synthetic seawater), $0.1\,\text{M}$ HCl (dilute
  acid), tap water (the baseline). Optional: a fourth solution that
  introduces a complication ($0.1\,\text{M}$ NH$_4$OH to provoke
  brass ammonia-cracking; $5\,\%$ NaCl plus aerated cycling to
  provoke pitting in stainless steel; deaerated solutions to test
  oxygen dependence).
\item Analytical balance reading to $\pm 0.1\,\text{mg}$. A digital
  jewellers' or chemistry-laboratory balance suffices.
\item Calipers, polishing paper ($600$ -- $1200$ grit), acetone for
  cleaning, paper towels.
\end{itemize}

\textbf{Procedure}.

\begin{itemize}[leftmargin=*]
\item Polish each coupon to a consistent finish, degrease with
  acetone, dry, weigh, record. Measure surface area to $\pm 1\,\%$.
\item Suspend each coupon in its assigned solution by an inert thread
  or wire (nylon line, not metal). The coupon should be fully
  submerged, not contacting the wall or another coupon.
\item Expose for a chosen interval: $48\,\text{h}$ for mild steel
  in HCl (fast reaction), $7$ -- $14\,\text{d}$ for the other
  combinations (slower reactions). Photograph the coupons at the start
  and end of exposure.
\item Remove, rinse in distilled water, gently brush off loose
  corrosion product, dry, weigh, record.
\item Compute the mass loss per unit area per unit time
  $\Delta m / (A \Delta t)$ and convert to corrosion rate
  $\text{rate} = \Delta m / (A \Delta t \rho)$ with $\rho$ the
  metal density. Report in $\text{mm}/\text{y}$ or $\text{mpy}$.
\end{itemize}

\textbf{Analysis}.

\begin{itemize}[leftmargin=*]
\item Tabulate the nine measured corrosion rates (three metals
  $\times$ three environments).
\item Identify the dominant corrosion mechanism in each combination
  (uniform attack, galvanic if two coupons were coupled, pitting,
  selective leaching) by visual inspection of the coupon surfaces.
\item Compare the measured rates with published corrosion data for
  these combinations (Fontana Table 4-1 or equivalent compilation).
  Discuss agreement and discrepancy.
\end{itemize}

\textbf{Reflection ($1{,}500$ to $2{,}500$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
\item Which metal-environment pair would the reader choose for a
  given engineering application (a heating-system tank, a marine
  fastener, an acid-handling vessel)?
\item What sources of measurement error dominated the experiment, and
  what would be done differently next time?
\item One mechanism the data revealed that the reader had not
  anticipated.
\end{itemize}

\textbf{Deliverable}. The mass and area measurements, the photographs
of coupons before and after exposure, the analysis, the report.

\textbf{Time}. Two to three hours of setup, the exposure interval as
above, two to three hours of measurement, four to six hours of
write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A mild-steel coupon ($50 \times 50 \times 1\,\text{mm}$, density
$7870\,\text{kg}/\text{m}^3$) loses $25\,\text{mg}$ in $7\,\text{d}$
in tap water. Compute the corrosion rate in $\text{mm}/\text{y}$.
\end{exercise}

\begin{exercise}
A galvanic couple of zinc and steel is immersed in $3\,\%$ NaCl. The
zinc-to-steel area ratio is $1:10$. The protective current is
$5\,\text{mA}$. Compute the corrosion rate of the zinc and the
expected service life of a $1\,\text{kg}$ zinc anode.
\end{exercise}

\begin{exercise}
A carbon-steel pipeline buried in soil of resistivity $30\,\si{\ohm}
\cdot\text{m}$ has external surface area $5{,}000\,\text{m}^2$. The
coating efficiency is $90\,\%$; the bare-steel current demand is
$15\,\text{mA}/\text{m}^2$. Compute the total protective current
required.
\end{exercise}

\begin{exercise}
A reinforced-concrete bridge deck has $50\,\text{mm}$ cover and an
effective chloride-diffusion coefficient $D = 1.5 \times 10^{-12}\,
\text{m}^2/\text{s}$. The threshold chloride concentration for
depassivation is $0.4\,\%$ by mass of cement and the surface
concentration in service is $1.5\,\%$. Estimate the initiation time
for corrosion of the reinforcing steel.
\end{exercise}

\begin{exercise}
An Archard wear analysis on a $50\,\text{N}$-loaded brake-pad-on-
disc contact gives a wear coefficient $K = 1 \times 10^{-3}$, a disc
hardness $H = 200\,\text{HV}$, and a sliding speed $v = 30\,\text{m}/
\text{s}$. Compute the volumetric wear rate.
\end{exercise}

\begin{exercise}
A pressurised-water-reactor pressure vessel accumulates $0.04\,\text{dpa}$
over $40\,\text{years}$ of service. The ductile-to-brittle-transition
shift is estimated as $\Delta T_{41J} = 40\,\degC$ for the steel
composition. Compute the upper-shelf reserve required at end of life.
\end{exercise}

\begin{exercise}
A polypropylene part is qualified to retain $50\,\%$ of its
elongation-at-break after $10{,}000\,\text{h}$ at $100\,\degC$. Using
the working rule that the time-to-property-loss doubles for every
$10\,\degC$ decrease in service temperature, estimate the service life
at $40\,\degC$.
\end{exercise}

\begin{exercise}
A type 316L stainless-steel piping system in a brewery cooling-water
service operates at $70\,\degC$ with $300\,\text{ppm}$ chloride. The
$K_{\text{ISCC}}$ for SCC in this combination is approximately
$10\,\text{MPa}\cdot\sqrt{\text{m}}$. A weld root has an estimated
residual stress of $200\,\text{MPa}$ and a crack-like flaw of
$2\,\text{mm}$. Assess SCC risk at this location.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive Faraday's law for metal dissolution and convert the corrosion
current density to a thickness-loss rate for iron, copper, and
aluminium.
\end{exercise}

\begin{exercise}
Derive the chloride-initiation time relation $t_{\text{init}} = c^2 /
(4 D [\text{erfc}^{-1}(C_{\text{th}}/C_s)]^2)$ from Fick's second law
with constant surface and zero initial chloride.
\end{exercise}

\begin{exercise}
Show that the Archard wear law follows from the assumption that
asperity-junction contacts shear at a fixed stress and that the
junction area is proportional to the applied load divided by the
hardness.
\end{exercise}

\begin{exercise}
Derive the working relation between dpa accumulated in a steel
exposed to a neutron flux $\Phi$ for a time $t$, given the
displacement cross-section $\sigma_d$ and the lattice atom density
$N_a$.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the rate of atmospheric corrosion of mild steel exposed
to a coastal industrial atmosphere. Compare with the rate in clean
rural atmosphere.
\end{exercise}

\begin{exercise}
Estimate the consequence of a $1\,\text{mm}$-deep pit on a
$3\,\text{mm}$-wall stainless-steel pipe carrying a flammable gas at
$5\,\text{MPa}$. State the residual safety factor.
\end{exercise}

\begin{exercise}
Estimate the wear life of a tyre tread, assuming a wear coefficient
$K = 10^{-4}$, a hardness $H = 5\,\text{MPa}$, and a typical highway
service profile.
\end{exercise}

\begin{exercise}
Estimate the time for UV embrittlement of an unstabilised
polyethylene rope in equatorial outdoor exposure. Compare with the
service life of the same rope under UV-stabilised formulation.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a one-dimensional finite-difference solver for chloride
diffusion into concrete cover with time-varying surface concentration.
Compute the initiation time for several cover thicknesses and compare
with the constant-surface analytical result.
\end{exercise}

\begin{exercise}
Simulate the polarisation curve of a metal under combined activation
and concentration polarisation, using the Butler-Volmer equation and
the Levich diffusion-limited current. Identify the corrosion potential
and corrosion current from the curve.
\end{exercise}

\begin{exercise}
Implement a Monte Carlo simulation of multiple-site fatigue-crack
growth at the rivet rows of a lap joint, with corrosion-accelerated
Paris-law constants. Compute the time to link-up failure as a function
of corrosion exposure.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify a corrosion-protection package for a buried steel oil-and-gas
pipeline in marshy ground (soil resistivity $\sim 20\,\si{\ohm}\cdot
\text{m}$). State the coating, the cathodic-protection method
(sacrificial or impressed-current), the design protective current
density, and the inspection regime.
\end{exercise}

\begin{exercise}
Design the cover-and-reinforcement specification for a reinforced-
concrete pier in a marine splash zone with a target service life of
$100\,\text{years}$. State the cover thickness, the concrete mix design,
the reinforcement type, and the inspection programme.
\end{exercise}

\begin{exercise}
Specify a wear-resistant surface for an excavator-bucket teeth set
that is currently wearing out at three-month replacement intervals.
State the substrate alloy, the surface treatment or coating, the
thickness, and the expected extension of replacement interval.
\end{exercise}

\begin{exercise}
Specify the materials selection for a chemical-plant heat-exchanger
shell carrying $80\,\degC$ saltwater on the shell side and process
fluid on the tube side. State the shell alloy, the tube alloy, the
gasket material, and the cathodic-protection or coating plan.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A type 304 stainless-steel cooling tower fan blade has cracked
through the trailing edge after eight years in service. The cracks
are branched and follow a path roughly perpendicular to the chord.
Diagnose the failure mode and recommend a single change to material
or operating environment that would prevent recurrence.
\end{exercise}

\begin{exercise}
A buried steel water main shows a pattern of pitting on its lower
half (the part in contact with the soil) and uniform attack on its
upper half (the part in the backfill stone). Diagnose the local
mechanism and propose an explanation for the asymmetry.
\end{exercise}

\begin{exercise}
A bicycle's high-strength aluminium frame fractures at the seat-post
clamp after three years of use. The fracture surface shows multiple
crack origins along the clamp gap and a clean fracture morphology
between them. Diagnose the failure mode and the contributing factors.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A swimming-pool installation has its type 316 stainless-steel ceiling
support fasteners failing one at a time after $15\,\text{years}$ of
service. The pool atmosphere is $30\,\degC$, $80\,\%$ relative humidity,
and contains $2\,\text{ppm}$ chloramine. Identify the failure mode and
the engineering remediation now standard in indoor-pool ceiling
specifications.
\end{exercise}

\begin{exercise}
A pressurised-water-reactor pressure-vessel mid-life surveillance
specimen shows a Charpy upper-shelf energy reduction of $20\,\%$
compared with the unirradiated baseline. Identify the controlling
material and operational parameters and the regulatory regime that
sets the life-extension decision.
\end{exercise}

\begin{exercise}
An offshore wind-turbine monopile coating system shows blistering and
isolated rust spots after three years of submerged service. Identify
the likely coating-failure modes and the inspection-and-repair
strategy that would extend the structure's service life.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A colleague proposes to specify a galvanic couple of stainless-steel
fasteners on a carbon-steel structure for a marine application,
arguing that the small fastener area-to-structure area ratio limits
the corrosion of the larger member. Write a one-paragraph reply
identifying the correct concern and the design adjustment required.
\end{exercise}

\begin{exercise}
A facility manager proposes to extend the inspection interval of a
chloride-stress-corrosion-cracking-prone vessel from one year to five
years on cost grounds. Write a one-paragraph evaluation of the
proposal, identifying the risk profile and the inspection alternatives.
\end{exercise}

\begin{exercise}
A start-up proposes a self-healing coating that repairs scratches and
holidays without external intervention. Write a one-paragraph reply
identifying the working questions about coverage, depth of healing,
and integration with the standard inspection regime.
\end{exercise}
