\chapter{Space infrastructure: orbital and planetary engineering}
\label{vol12:ch10}

\epigraph{In space, the constraints are stricter, the failure modes are sharper, the feedback is slower, and the engineering discipline shows.}{Author}

\chapmeta{Half-life: medium-life. Archetypes: scaling, control, failure. Exercise target: 22. Project track: design (CubeSat-class mission specification, one-week effort, no build). Hazard class: high (multi-billion-dollar single-event failure exposure; orbital-debris implications). Reader path default: standard, with the orbital-mechanics and constellation-engineering sections core and the crewed-spaceflight and Mars-and-lunar sections mastery. Named cases: Hubble Space Telescope spherical-aberration mirror error (\cite{acc:nasa-hubble-optical-systems-1990}); Mars Climate Orbiter unit-conversion failure (\cite{acc:nasa-mco-mib}); Iridium-Cosmos 2009 collision (\cite{paper:weeden-iridium-cosmos-2010}).}

\noindent
Space is the layer of civilization infrastructure now sitting above the atmosphere. Approximately $10{,}000$ active satellites in orbit as of 2024 (up from $\approx 1{,}000$ in 2018); approximately $100$ launches/yr globally and rising; approximately $\$500$ billion annual global space economy (including downstream services); approximately $80{,}000$ tracked debris objects $>10\,\text{cm}$ in low-Earth orbit; the GPS, Galileo, GLONASS, and BeiDou positioning systems that the contemporary civilian economy depends on; the weather-and-Earth-observation satellites that the climate, agriculture, water, and disaster-management infrastructures of every country depend on; the satellite-communications constellations (Starlink at $\approx 6{,}000$, Iridium, OneWeb, others) that are restructuring the global communications infrastructure. The chapter takes the engineering of this layer as the topic, with the orbital-debris and sustainability question as the failure mode.

\section{Orbital mechanics for engineers}\pathtag{core}

Orbital mechanics is the engineering application of Kepler's laws and Newton's gravitation to the motion of objects in space. The key parameters: semi-major axis $a$, eccentricity $e$, inclination $i$, longitude of ascending node $\Omega$, argument of periapsis $\omega$, true anomaly $\nu$. The orbital period $T = 2\pi\sqrt{a^3/\mu}$ where $\mu = GM_{\text{Earth}} = 3.986 \times 10^{14}\,\text{m}^3/\text{s}^2$. For Earth orbits: $T = 90\,\text{min}$ at $\approx 400\,\text{km}$ altitude (LEO); $T = 12\,\text{h}$ at $\approx 20{,}000\,\text{km}$ (MEO; GPS region); $T = 24\,\text{h}$ at $\approx 36{,}000\,\text{km}$ altitude (GEO).

The orbital choice determines the mission profile. LEO (low-Earth orbit; $\sim 200$-$2{,}000\,\text{km}$) offers low launch cost, high spatial resolution for imaging, low communication latency, but each satellite covers only a small ground footprint at any time and the orbit decays over $\sim 5$-$25$ years from atmospheric drag. MEO (medium-Earth orbit; $\sim 2{,}000$-$36{,}000\,\text{km}$) offers larger ground coverage and is used by the global-navigation-satellite-system constellations. GEO (geostationary; $35{,}786\,\text{km}$ altitude over the equator) offers a fixed sub-satellite point and is used by broadcast television, weather satellites, and some communications; the GEO ring is a finite resource regulated by the ITU. SSO (Sun-synchronous orbit; $\sim 600$-$800\,\text{km}$ with specific inclination) maintains a fixed local-solar-time at each ground location and is used for Earth observation.

\begin{archetype}[Control: orbital stationkeeping]
A satellite in orbit experiences perturbations: atmospheric drag (LEO; substantial); Earth's oblateness ($J_2$ secular drift in inclination and longitude of ascending node; all orbits); third-body gravitation (Moon, Sun; substantial at GEO); solar radiation pressure (substantial for large surface-to-mass ratio satellites). The satellite must periodically execute stationkeeping manoeuvres to maintain its designed orbit. For a GEO satellite the annual stationkeeping budget is $\approx 50\,\text{m/s}$ of $\Delta v$; over a $15$-year mission, $\approx 750\,\text{m/s}$ of $\Delta v$; the propellant capacity is sized for this plus end-of-life disposal. The mission-end is typically when stationkeeping propellant runs out and the satellite drifts to a graveyard orbit (GEO+$300\,\text{km}$ per IADC guidelines).
\end{archetype}

\begin{estimation}
\textbf{Question.} Estimate the $\Delta v$ required to launch a $1{,}000\,\text{kg}$ satellite to LEO, GEO, and lunar-transfer-orbit.

\textbf{Reasoning.} Velocity at LEO $\approx 7.8\,\text{km/s}$; surface velocity (rotation; equator) $\approx 0.46\,\text{km/s}$; required launch $\Delta v$ to LEO $\approx 9.5\,\text{km/s}$ including gravity and drag losses. GEO requires LEO insertion ($9.5\,\text{km/s}$) plus Hohmann transfer ($\approx 2.4\,\text{km/s}$ apogee burn) plus circularisation at GEO ($\approx 1.5\,\text{km/s}$); total $\approx 13.4\,\text{km/s}$. Lunar transfer requires LEO ($9.5\,\text{km/s}$) plus trans-lunar injection ($\approx 3.1\,\text{km/s}$) plus lunar-orbit insertion ($\approx 0.9\,\text{km/s}$); total $\approx 13.5\,\text{km/s}$. Conclusion: $\Delta v$ requirements increase by $\sim 40\%$ from LEO to GEO or lunar; the corresponding propellant fraction (rocket equation) makes lunar missions substantially more expensive than LEO missions and explains the economic and engineering case for in-space propellant depots and reusable stages.
\end{estimation}

\section{Launch systems and propulsion}\pathtag{standard}

The Tsiolkovsky rocket equation $\Delta v = I_{sp} g_0 \ln(m_0/m_f)$ is the foundational constraint of launch system design. $I_{sp}$ (specific impulse, in seconds) characterises the propulsion efficiency: $\sim 280\,\text{s}$ for solid rocket motors; $\sim 320\,\text{s}$ for kerosene-LOX (RP-1/LOX, e.g., Falcon 9 first stage); $\sim 360\,\text{s}$ for methane-LOX (Raptor, BE-4); $\sim 450\,\text{s}$ for hydrogen-LOX (RS-25, RL-10, Vulcain). For a single-stage launch to LEO, the propellant mass fraction required is $\sim 0.95$, which is at the structural limit of any current technology and explains why all operational launch systems are multi-stage.

The contemporary launch-system economics has been substantially reshaped by reusable first stages (SpaceX Falcon 9; the projected reusability of New Glenn, Vulcan, Starship). Falcon 9 launch cost to LEO has dropped from $\sim \$10{,}000$/kg in the early 2010s to $\sim \$3{,}000$/kg in the late 2010s under reusability; Starship (when operational at design specifications) is projected to reach $<\$500$/kg. The reduction has enabled the LEO-constellation deployments (Starlink at $\sim 6{,}000$ satellites and growing) and is reshaping the engineering economics of space missions broadly. The downstream implications: more satellites, more debris (next section); more capability per dollar; a wider range of feasible missions.

\section{Satellites: design, lifetime, end-of-life}\pathtag{standard}

A satellite's design integrates structure (mechanical loads of launch; thermal cycling in orbit), thermal control (radiative balance against direct and Earth-reflected solar input), power (solar arrays sized for end-of-life; batteries for eclipse periods; power distribution), attitude-and-orbit-control (gyros, reaction wheels, magnetorquers, thrusters; sensors including sun, star, Earth, GPS; control algorithms), propulsion (chemical for large $\Delta v$; electric for efficient long-duration manoeuvres), telemetry, tracking, and command (TT\&C), and the payload (the mission-specific instrumentation). The integration is the spacecraft-systems-engineering discipline.

Satellite lifetimes range from months (for some experimental CubeSats) to $15$-$20$ years for major GEO communications satellites. The end-of-life challenge: a satellite that fails or runs out of propellant in operational orbit becomes long-lived debris. The IADC (Inter-Agency Space Debris Coordination Committee) guidelines call for LEO satellites to deorbit within $25$ years of mission end; GEO satellites to manoeuvre to a graveyard orbit. Compliance is good for state and major commercial operators but remains a regulatory and operational challenge for smaller and developing-country operators.

\section{Constellation engineering: Starlink, Galileo, GPS, OneWeb}\pathtag{core}

A constellation is a coordinated group of satellites designed to provide a service that no single satellite can. The major examples include: GPS (24 operational satellites in $6$ orbital planes at $20{,}200\,\text{km}$ altitude, $55^{\circ}$ inclination, providing global positioning with $\sim 3\,\text{m}$ accuracy); Galileo, GLONASS, BeiDou (comparable global-navigation satellite systems); Iridium (66 satellites in $6$ orbital planes at $780\,\text{km}$, providing global voice and data communications); Starlink ($\sim 6{,}000$ satellites at $\sim 550\,\text{km}$, providing global broadband internet); OneWeb ($\sim 650$ satellites at $\sim 1{,}200\,\text{km}$, similar role to Starlink); the projected Project Kuiper, Guowang, and others. Each constellation's design balances the global-coverage requirement, the per-user-data-rate requirement, the launch and operational cost, and the regulatory constraints.

The contemporary LEO-constellation engineering is at the frontier of satellite mass-production and deployment. SpaceX's Starlink production line is reportedly $\sim 100$ satellites/month; launch cadence at $\sim 60$ satellites/launch via Falcon 9 ride-share. The engineering implications: high-volume satellite production with industrial-manufacturing standards (rather than the prior craft-built production); automated deployment, station-keeping, collision-avoidance, and end-of-life management; aggregate constellation-level systems engineering that exceeds the per-satellite engineering in complexity. The civilian-and-commercial impact: broadband internet to previously unserved populations; competitive pressure on terrestrial mobile broadband; new platform for downstream services (in-flight Wi-Fi; maritime connectivity; remote-region connectivity).

\section{Communications and earth observation}\pathtag{standard}

Earth observation from space supports the weather forecasting, climate monitoring, disaster response, agricultural monitoring, urban planning, and environmental monitoring infrastructures that nearly every other chapter in this volume depends on. The major civilian constellations: the geostationary weather satellites (GOES, Meteosat, Himawari, FengYun, COMS); the polar-orbiting weather satellites (NOAA, MetOp, FengYun-3); the high-resolution commercial-imagery satellites (Maxar, Planet Labs, Capella Space, ICEYE); the open-access programmes (Landsat since 1972, Sentinel since 2014 under the EU Copernicus programme); the specialised platforms (GRACE for gravity; ICESat for ice; Aeolus for wind; CHAMP, SWARM for magnetic field; the constellation of climate-monitoring satellites). The instrument portfolio: optical imaging (panchromatic, multispectral, hyperspectral); synthetic-aperture radar; passive microwave; LiDAR; gravimeters; magnetometers.

Communications satellites have been the historical mainstay of the commercial space industry. The traditional geostationary satellites continue to operate (Intelsat, SES, Eutelsat, Telesat, others); the contemporary LEO constellations are restructuring the market. The downstream services: direct-to-home television; mobile-satellite voice and data; maritime communications; in-flight Wi-Fi; emerging direct-to-handset services (Starlink-direct, AST SpaceMobile).

\section{Crewed spaceflight and space stations}\pathtag{mastery}

Crewed spaceflight is the engineering of life-support and crew operations in space. The current operational systems: the International Space Station ($\sim 420{,}000\,\text{kg}$ mass; $\sim 380\,\text{km}$ altitude; continuously crewed since 2000; planned deorbit $\sim 2030$); the Chinese Tiangong space station; the commercial crewed missions (SpaceX Crew Dragon to ISS; Polaris Dawn; the projected Axiom modules and the Vast Space station). The engineering: life-support (oxygen generation, CO$_2$ removal, water recycling, food, waste management, thermal control, radiation shielding); crew-support (sleep stations, exercise equipment, medical, food preparation); operations (logistics resupply; crew rotation; experiments; maintenance; contingency operations). The international cooperation around the ISS (NASA, Roscosmos, ESA, JAXA, CSA) is one of the longer-running and more substantial international engineering collaborations of the contemporary period.

\section{Mars and lunar engineering: Artemis, ISRU}\pathtag{mastery}

The contemporary lunar return is the Artemis programme (NASA; with international and commercial partners; first crewed landing originally planned 2024 then 2026, now likely later). The engineering scope: the Space Launch System (the heavy-lift rocket for Orion); the Orion crew capsule; the Human Landing System (SpaceX Starship variant; Blue Origin's Blue Moon as second provider); the Gateway lunar orbital station; the surface infrastructure for sustained presence. The Chinese lunar programme is on a parallel trajectory with a different architecture and a partly-different timeline.

The Mars engineering is at the longer-term planning stage. The technical challenges: the $\Delta v$ requirements for crewed Mars mission ($\sim 16$-$20\,\text{km/s}$ outbound; entry-descent-landing of the substantial mass required for a crewed mission; return propellant manufacture from local resources). In-situ resource utilisation (ISRU) is the engineering of using local resources to reduce mass-from-Earth requirements: water-ice extraction (Moon and Mars); oxygen production from regolith or CO$_2$; propellant manufacture from local hydrogen and oxygen or methane and oxygen sources. The ISRU engineering is at the technology-demonstration stage and is a key enabler for sustained crewed presence beyond low-Earth orbit.

\section{Space debris and orbital sustainability}\pathtag{core}

Space debris is the orbital legacy of $65$ years of space activity. The current debris environment: $\sim 36{,}000$ tracked objects $>10\,\text{cm}$ in size; $\sim 10^6$ objects $1$-$10\,\text{cm}$ (untracked but capable of catastrophic damage); $\sim 10^8$ objects $1\,\text{mm}$-$1\,\text{cm}$. The Iridium-Cosmos collision of 2009 \cite{paper:weeden-iridium-cosmos-2010} (active Iridium-33 communications satellite collided with defunct Russian Cosmos-2251; both destroyed; both created $\sim 1{,}500$ trackable debris objects each, more than doubling the population at the $\sim 800\,\text{km}$ altitude band) and the Chinese Fengyun-1C ASAT test of 2007 (deliberate destruction; $\sim 3{,}000$ trackable debris objects) together account for substantial fractions of the contemporary trackable-debris population.

The Kessler syndrome (Kessler and Cour-Palais 1978; \cite{paper:kessler-1978}) is the potential failure mode: a runaway debris-on-debris collision cascade that renders certain orbital regions unusable for decades. The contemporary engineering response: post-mission disposal (the IADC 25-year guideline for LEO; graveyard orbit for GEO); pre-mission collision-avoidance planning; on-orbit conjunction-assessment and avoidance manoeuvres; active debris removal demonstrators (ClearSpace-1; Astroscale ELSA); regulatory action (the FCC's revised 5-year rather than 25-year deorbit requirement; the UN COPUOS Long-Term Sustainability guidelines). The constellations of $10^4$ satellites projected for 2030 stress all aspects of this regime and make orbital-sustainability one of the central contemporary engineering-and-policy questions in space.

\begin{principle}[Orbital sustainability as engineered scope]
Orbital sustainability is the engineered property that future generations have access to the same orbital regimes that current operators do. The engineering elements: per-mission post-mission disposal compliance; pre-mission collision-assessment as standard practice; on-orbit conjunction-avoidance capability; active-debris-removal as an emerging capability. The institutional elements: international guidelines under UN COPUOS; national-licensing regulatory enforcement; the responsibility of constellation operators for the cumulative impact of their fleets. The engineering profession's responsibility includes the per-mission compliance and the engagement with the institutional structures that will determine whether the contemporary expansion of orbital activity is sustainable across the multi-decade horizon.
\end{principle}

\section{Failure: launch failures, in-orbit losses, mission failures}\pathtag{core}

The chapter's failure cases each illustrate a distinct mode of space-systems engineering failure.

The Hubble Space Telescope spherical-aberration mirror error \cite{acc:nasa-hubble-optical-systems-1990} is the case of fabrication-and-quality-assurance failure: the primary mirror was ground to high precision but to the wrong figure because of a misaligned null-corrector during the metrology; the error was detectable in ground tests but the alternative tests that would have detected it were skipped for cost reasons; the post-launch correction required the COSTAR optical-correction package installed by Servicing Mission 1 in 1993 and built into all subsequent instruments. The engineering lesson: the cross-checks of quality-critical parameters by independent methods should not be skipped under cost pressure.

The Mars Climate Orbiter unit-conversion failure of 1999 \cite{acc:nasa-mco-mib} is the case of interface-control failure: the trajectory-computation software used pound-seconds (imperial) for thrust impulse while the navigation team expected newton-seconds (metric); the resulting orbit-insertion was much lower than planned and the spacecraft was destroyed in the Martian atmosphere. The engineering lesson: interface-control discipline (the documented agreement on units, formats, conventions across organisational boundaries) is critical and was substantially under-applied. The case is widely taught in software and systems engineering as the iconic interface-control failure.

The Iridium-Cosmos collision \cite{paper:weeden-iridium-cosmos-2010} is the case of conjunction-assessment failure: the operational status of the Cosmos satellite and the conjunction-prediction accuracy were both insufficient to identify the collision risk; the resulting debris cloud increased the debris population in the affected altitude band by $\sim 25\%$ overnight. The engineering response: the substantial improvement of conjunction-assessment capability (the US Space Force's $18$-second-period catalog; commercial services such as LeoLabs); the operational practice of routine conjunction-avoidance manoeuvres for active satellites. The longer-term lesson: as orbital populations grow, the conjunction-management infrastructure must grow correspondingly.

\section{Project}\pathtag{core}

\begin{project}[CubeSat-class mission specification]
Specify a CubeSat-class mission (3U-12U) end to end. Document: the mission objective; the payload; the orbit selection (altitude, inclination, lifetime, launch availability); the bus design (mass, power, attitude control, communications, propulsion if any); the link budget (uplink and downlink; modulation; bandwidth; ground station); the operations concept (commanding, telemetry processing, payload-data processing); the end-of-life plan (deorbit mechanism; compliance with the 25-year or 5-year guideline as applicable). The project deliverable is the mission-specification document; the engineering analysis is the trade-off between mission capability, mission cost, and orbital-sustainability compliance.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A satellite is in a circular orbit at $400\,\text{km}$ altitude. Compute the orbital velocity, the orbital period, and the orbital angular velocity.
\end{exercise}

\begin{exercise}
A geostationary satellite at $35{,}786\,\text{km}$ altitude has a $15$-year mission life with $50\,\text{m/s}$ annual stationkeeping budget plus $200\,\text{m/s}$ for end-of-life graveyard orbit. Compute the total $\Delta v$ budget and the propellant mass required for a $3{,}000$-kg satellite with $300$-s $I_{sp}$ thrusters.
\end{exercise}

\begin{exercise}
A LEO constellation has $1{,}500$ satellites at $550\,\text{km}$ altitude, each $250\,\text{kg}$ mass. Compute the orbital decay time at this altitude (typical $\sim 5$ years), the launch-cost-per-year for replacement at $\$3{,}000$/kg, and the annual launch cadence required for steady-state.
\end{exercise}

\begin{exercise}
A communications-satellite downlink at $20\,\text{GHz}$ with $10\,\text{W}$ transmit power, $40\,\text{dB}$ antenna gain, propagating $40{,}000\,\text{km}$ to a ground antenna with $50\,\text{dB}$ gain. Compute the received power and the signal-to-noise ratio at a noise temperature of $200\,\text{K}$.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the Tsiolkovsky rocket equation from conservation of momentum. Identify the propellant mass fraction required for a single-stage-to-orbit launch.
\end{exercise}

\begin{exercise}
Derive the Hohmann-transfer $\Delta v$ from a circular orbit of radius $r_1$ to a circular orbit of radius $r_2$.
\end{exercise}

\begin{exercise}
Derive the orbital-decay rate from atmospheric drag as a function of altitude and ballistic coefficient. Identify the typical decay times at $400\,\text{km}$, $600\,\text{km}$, and $800\,\text{km}$ altitudes.
\end{exercise}

\begin{exercise}
Derive the $J_2$ secular drift of orbital plane (precession of the ascending node) and identify the inclination for sun-synchronous orbit at a given altitude.
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the total launch-cost-per-kg reduction implied by reusable launch vehicles relative to expendable, given the relative production-to-refurbishment cost ratios.
\end{exercise}

\begin{exercise}
Estimate the orbital-debris collision risk for a satellite in a popular LEO orbit, given current debris density and the satellite cross-section. Identify the implications for conjunction-avoidance-manoeuvre frequency.
\end{exercise}

\begin{exercise}
Estimate the in-orbit servicing market for refuelling and life-extension of GEO satellites at current pricing and projected operator demand.
\end{exercise}

\begin{exercise}
Estimate the propellant production rate required from a lunar in-situ resource utilisation plant to support a sustained crewed lunar surface programme.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Simulate a Kessler-cascade scenario starting from current orbital-debris distribution. Identify the constellation-density thresholds and active-debris-removal rates that prevent versus enable the cascade.
\end{exercise}

\begin{exercise}
Simulate a global-positioning-system constellation outage scenario. Identify the cross-civilian-economy impact (aviation, maritime, banking timing, communications synchronisation).
\end{exercise}

\begin{exercise}
Simulate the link-budget for a LEO satellite-to-ground link under fading and weather conditions. Identify the margins required for service-availability targets.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design an Earth-observation mission for a specific monitoring objective (e.g., methane-plume detection from oil-and-gas infrastructure; deforestation monitoring; sea-level rise). Specify orbit, payload, revisit, ground processing.
\end{exercise}

\begin{exercise}
Design a constellation-management system for a $1{,}000$-satellite LEO constellation including collision-avoidance, end-of-life disposal, and operational availability monitoring.
\end{exercise}

\begin{exercise}
Design an active-debris-removal mission targeting a defunct large object in LEO. Specify the rendezvous strategy, the capture mechanism, the deorbit propulsion, the cost and timeline.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Discuss the Mars Climate Orbiter unit-conversion failure as an interface-control failure. Identify the engineering and organisational decisions that contributed; identify the structural improvements that would prevent recurrence.
\end{exercise}

\begin{exercise}
Argue: orbital sustainability is the responsibility of the engineering profession; the operators of large LEO constellations are accountable for the cumulative effect of their fleets on the orbital environment. Argue the opposite. Identify the implications for constellation-engineering practice and for international space governance.
\end{exercise}

\begin{exercise}
Discuss the contemporary expansion of LEO commercial constellations. Identify the engineering and policy implications for orbital-debris management; identify the regulatory frameworks required for sustained orbital activity.
\end{exercise}

\begin{exercise}
Discuss the Hubble Space Telescope mirror failure as a quality-assurance failure. Identify the lessons for high-stakes single-event engineering and for the cost-pressure-on-quality dynamics.
\end{exercise}
