\chapter{Vibrations}
\label{vol03:ch06}

\epigraph{The problem of vibrating systems, although it may at first
sight appear to be of a special character, has a peculiar importance
arising from its application to many of the most refined parts of
physical inquiry.}{Lord Rayleigh, \emph{The Theory of Sound}, vol.~I
(1877), preface \cite[p.~v]{hist:rayleigh1877}.}

\chapmeta{Half-life: durable. Archetypes: stability, control, failure. Exercise target: 35. Project track: physical measurement plus analysis (hybrid). Hazard class: none. Reader path default: standard, with sections explicitly tagged. Named cases: Tacoma Narrows Bridge (1940), London Millennium Footbridge (2000).}

\noindent
A mechanical system that returns toward an equilibrium when displaced
oscillates around that equilibrium. The oscillation is the system's
vibration. Vibrations carry energy, transmit force, radiate sound,
fatigue materials, and at the wrong frequency, destroy what they act
on. The chapter develops the engineering theory of vibration: the
oscillator's natural frequency, its damping, its forced response, its
modal structure as the system grows from one degree of freedom to
many to a continuum, and the diagnostic discipline that turns
vibration measurements into engineering decisions.

\section{Single-degree-of-freedom oscillators}\pathtag{core}

The simplest vibrating system is a mass on a spring. A mass $m$
connected to a wall by a linear spring of stiffness $k$, on a
frictionless horizontal surface, has displacement $x(t)$ measured
from the spring's equilibrium length. Newton's second law (or the
Lagrangian formulation of chapter 5) gives the equation of motion
\[
m\ddot{x} + kx = 0.
\]
The equation is linear, homogeneous, and second-order. Its general
solution is
\[
x(t) = A\cos(\omega_n t) + B\sin(\omega_n t) = X\cos(\omega_n t -
\phi),
\]
where
\[
\omega_n = \sqrt{\frac{k}{m}}
\]
is the \engterm{natural angular frequency}, and the constants $A,
B$ (or equivalently $X, \phi$) are set by the initial conditions
$x(0)$ and $\dot{x}(0)$. The natural frequency is the rate at which
the system oscillates when displaced from equilibrium and left to
itself; it depends only on the stiffness and the mass, not on the
amplitude of motion. The natural frequency in Hz is $f_n = \omega_n /
(2\pi)$, and the natural period is $T_n = 1/f_n = 2\pi/\omega_n$.

The energy of the oscillator is
\[
E = \tfrac{1}{2} m \dot{x}^2 + \tfrac{1}{2} k x^2,
\]
which is constant in time for the undamped system. The kinetic
energy is maximum at the equilibrium position ($x = 0$, $\dot{x} =
\pm X\omega_n$); the potential energy is maximum at the extremes ($x
= \pm X$, $\dot{x} = 0$). Energy oscillates between the two forms
at frequency $2 \omega_n$ (twice the displacement frequency, because
each form peaks once per quarter-period).

Many real systems reduce to the single-degree-of-freedom (SDOF)
oscillator in their simplest analysis. A pendulum of length $L$
linearised at small amplitude has $\omega_n = \sqrt{g/L}$. A rigid
body of mass moment of inertia $I$ on a torsional spring of
stiffness $K$ has $\omega_n = \sqrt{K/I}$. A floor joist supporting
a load $m$ at midspan has, for small vertical vibrations, $\omega_n
= \sqrt{k_{\text{joist}}/m}$ with $k_{\text{joist}} = 48 EI/L^3$
the joist's flexural stiffness at midspan. The same equation, with
different physical interpretations of $m$ and $k$, recurs across
domains.

\engterm{Equivalent stiffness and equivalent mass}. The reduction of
a complex structure to a single equivalent SDOF system is the first
step in many vibration analyses. For a long beam vibrating in its
fundamental mode, the equivalent mass is not the full beam mass but
the modal-effective fraction (typically $0.5$ to $0.6$ for common
boundary conditions); the equivalent stiffness is the modal
stiffness corresponding to that mass. The same approach reduces a
multi-storey building's lowest sway mode to an SDOF system, a
spinning rotor's first whirl mode to an SDOF system, or a complex
machine's principal vibration mode to an SDOF system. The reduction
is exact only for the mode it represents; other modes need their
own reductions.

\section{Damped vibration}\pathtag{core}

Real systems lose energy to friction, internal-material hysteresis,
air resistance, and acoustic radiation. The simplest model adds a
linear viscous damper of damping coefficient $c$:
\[
m\ddot{x} + c\dot{x} + kx = 0.
\]
The equation is still linear; its solutions depend on the damping
strength relative to the system's natural frequency. The
\engterm{damping ratio} is
\[
\zeta = \frac{c}{2\sqrt{mk}} = \frac{c}{2 m \omega_n},
\]
a dimensionless quantity. Three regimes:

\engterm{Underdamped} ($\zeta < 1$): the system oscillates with
exponentially decaying amplitude. The solution is
\[
x(t) = X_0 e^{-\zeta\omega_n t} \cos(\omega_d t - \phi),
\]
where
\[
\omega_d = \omega_n\sqrt{1 - \zeta^2}
\]
is the damped angular frequency. For lightly damped systems
($\zeta \lesssim 0.1$), $\omega_d \approx \omega_n$, and the period
is approximately the undamped period. Each successive amplitude
peak is reduced from the previous one by the factor $e^{-2\pi\zeta/
\sqrt{1-\zeta^2}}$; the logarithm of this ratio is the
\engterm{logarithmic decrement} $\delta = 2\pi\zeta/\sqrt{1-\zeta^2}$,
a standard experimental measure of damping
(Figure~\ref{fig:vol03ch06:free-decay}).

\input{volumes/03-force/ch06-vibrations/figures/fig-free-decay.tex}

The logarithmic decrement is the cleanest single-number reading for
light damping. Mark two amplitude peaks $n$ cycles apart, take the
ratio $x_1/x_{1+n}$, and the per-cycle decrement is $\delta =
(1/n)\ln(x_1/x_{1+n})$. Inverting the definition gives $\zeta =
\delta/\sqrt{4\pi^2 + \delta^2}$, which for the common light-damping
case $\delta \ll 1$ reduces to $\zeta \approx \delta/(2\pi)$. A
structure whose free vibration halves in amplitude every ten cycles
has $\delta = (1/10)\ln 2 = 0.069$ and $\zeta \approx 0.011$, in the
range of a bolted-steel frame. The companion program
\path{code/sdof_response.py} integrates the free response numerically
and recovers $\zeta$ from the simulated peaks, reproducing the input
damping ratio to within the peak-resolution limit.

\engterm{Critically damped} ($\zeta = 1$): the system returns to
equilibrium in the shortest time without oscillating. The solution
is
\[
x(t) = (A + Bt) e^{-\omega_n t}.
\]
Critical damping is the dividing line between oscillatory and
non-oscillatory return; many engineering systems (vehicle shock
absorbers, instrument needles, door closers) are designed near
critical damping.

\engterm{Overdamped} ($\zeta > 1$): the system returns to equilibrium
without oscillating, but more slowly than at critical damping. The
solution is a sum of two decaying exponentials with different rates.
Overdamped systems are sluggish; they appear in heavily-damped
isolators and in some control-loop responses.

Typical damping ratios for engineering systems, current as of 2025
per the standard structural-dynamics references \cite{text:den-hartog-vibrations}:

\begin{itemize}[leftmargin=*]
  \item Welded steel structures: $\zeta \approx 0.005$ to $0.01$.
  \item Bolted steel structures: $\zeta \approx 0.01$ to $0.02$.
  \item Reinforced concrete buildings: $\zeta \approx 0.02$ to $0.05$.
  \item Soils under cyclic loading: $\zeta \approx 0.05$ to $0.20$.
  \item Vehicle suspensions (designed): $\zeta \approx 0.2$ to $0.7$.
  \item Automotive shock absorbers (alone, in compression): $\zeta
    \approx 0.3$.
  \item Pendulum clock: $\zeta \approx 10^{-4}$ (very lightly damped,
    by design).
\end{itemize}

The numerical values vary widely by structure, material, joint
detailing, amplitude, and frequency; published values are working
approximations, and a specific design uses the measured damping of
the actual structure when it matters. The reference values above are
collected in machine-readable form in \path{data/damping_table.csv}
with the dominant mechanism noted for each structure type.

\engterm{Sources of damping}. Viscous damping is the simplest
mathematical model and represents only a few real dissipation
mechanisms (motion through a fluid at low Reynolds number, properly
designed dashpots). Real damping in structures is dominated by
material hysteresis, frictional sliding at joints, and acoustic
radiation; these mechanisms are nonlinear and amplitude-dependent in
detail but are commonly modelled as equivalent viscous damping for
engineering analysis. The reduction is accurate near resonance for
small amplitudes and increasingly approximate elsewhere.

\section{Forced response and resonance}\pathtag{core}

A vibrating system subjected to a time-varying applied force
$F(t)$ has the equation of motion
\[
m\ddot{x} + c\dot{x} + kx = F(t).
\]
The general solution is the sum of the homogeneous (free) response,
which decays in time for $\zeta > 0$, and a particular (steady-state)
response driven by $F(t)$. The free response is determined by initial
conditions and dies out; the steady-state response continues
indefinitely so long as $F(t)$ is applied.

For a sinusoidal driving force $F(t) = F_0 \cos(\omega t)$, the
steady-state response is
\[
x_{\text{ss}}(t) = X(\omega) \cos(\omega t - \phi(\omega)),
\]
where the response amplitude and phase are
\[
X(\omega) = \frac{F_0/k}{\sqrt{(1 - r^2)^2 + (2\zeta r)^2}}, \quad
\tan\phi(\omega) = \frac{2\zeta r}{1 - r^2},
\]
with $r = \omega/\omega_n$ the dimensionless frequency ratio. The
quantity $F_0/k$ is the static deflection the spring would
experience under a constant force $F_0$; the ratio $X/(F_0/k)$ is
the \engterm{dynamic amplification factor}.

At $\omega \ll \omega_n$ ($r \ll 1$), the amplification is
approximately $1$: the system follows the driving force quasi-
statically. At $\omega \gg \omega_n$ ($r \gg 1$), the amplification
approaches $1/r^2$: the system's mass-inertia dominates, and the
response amplitude is small compared to the static deflection. At
$\omega = \omega_n$ ($r = 1$), \engterm{resonance}, the amplification
is $1/(2\zeta)$, which can be very large for lightly damped systems.
A structure with $\zeta = 0.01$ at resonance has an amplification of
$50$; with $\zeta = 0.001$, it is $500$
(Figure~\ref{fig:vol03ch06:frequency-response}).

\input{volumes/03-force/ch06-vibrations/figures/fig-frequency-response.tex}

The peak of the amplification curve sits slightly below $r = 1$, at
$r_{\text{peak}} = \sqrt{1 - 2\zeta^2}$, and its height is
$1/(2\zeta\sqrt{1-\zeta^2})$, which for light damping is
indistinguishable from the resonant value $1/(2\zeta)$ at $r = 1$.
The resonant amplification factor $1/(2\zeta)$ is identical to the
\engterm{quality factor} $Q$ of the resonance, the same $Q$ that
appears in the analysis of electrical resonant circuits, optical
cavities, and atomic clocks: a high-$Q$ system stores many cycles'
worth of energy before dissipating it and responds sharply to a
narrow band of frequencies around $\omega_n$.

The resonance amplification is the working tool and the working
hazard of structural dynamics. A lightly damped natural frequency
at the frequency of a driving force amplifies the system's response
by $1/(2\zeta)$, which can exceed the structure's capacity even when
the static force is small. A driving force at any frequency well
away from $\omega_n$ produces a response no larger than $F_0/k$,
the static deflection. The design discipline is to either keep the
driving frequencies well separated from the natural frequencies, or
to add damping to limit the resonance amplification.

The phase $\phi$ has a corresponding pattern. At $\omega \ll
\omega_n$, the response is in phase with the force; at $\omega =
\omega_n$, the response lags by $90^{\circ}$ (the force is in phase
with the velocity, the displacement lags); at $\omega \gg \omega_n$,
the response lags by $180^{\circ}$ (the response is out of phase
with the force). The phase shift at resonance is the signature that
identifies a measured resonance experimentally: the displacement
and velocity are $90^{\circ}$ apart, and the driving force is in
phase with the velocity (Figure~\ref{fig:vol03ch06:phase}).

\input{volumes/03-force/ch06-vibrations/figures/fig-phase.tex}

The phase reading is more reliable than the amplitude reading for
locating a natural frequency. The amplitude peak moves with damping
(to $r_{\text{peak}} = \sqrt{1-2\zeta^2}$) and can be masked by noise
or by nearby modes; the $90^{\circ}$ phase crossing sits at exactly
$r = 1$ for every damping ratio. Experimental modal analysis uses the
phase crossing, not the amplitude peak, to fix $\omega_n$. A sampled
sine-sweep frequency response with both amplitude and phase columns
is provided in \path{data/sweep_response.csv} for a system with
$f_n = 24\,\si{\hertz}$ and $\zeta = 0.04$; reading the file confirms
that the phase passes through $90^{\circ}$ at $24\,\si{\hertz}$ while
the amplitude peak sits a fraction of a hertz lower.

\engterm{Beating near resonance}. When a lightly damped system is
driven close to but not exactly at its natural frequency, the
transient free response at $\omega_n$ and the steady-state response at
$\omega$ coexist before the transient decays. Two sinusoids at nearby
frequencies add to a fast oscillation at the average frequency inside
a slow envelope at the difference frequency $|\omega - \omega_n|$, a
beat (Figure~\ref{fig:vol03ch06:beat}). The beat period
$2\pi/|\omega-\omega_n|$ grows without bound as the drive approaches
exact resonance. For an undamped system driven exactly at $\omega_n$,
the response amplitude grows linearly in time rather than beating, the
mathematical signature of unbounded resonant buildup. Audible beats
from a running machine are the working warning that an excitation has
drifted close to a structural natural frequency.

\input{volumes/03-force/ch06-vibrations/figures/fig-beat.tex}

\engterm{Transmissibility}. A vibrating support transmits its
motion to a mounted mass. The dimensionless ratio of the mass's
amplitude to the support's amplitude is the transmissibility:
\[
\text{TR}(\omega) = \sqrt{\frac{1 + (2\zeta r)^2}{(1 - r^2)^2 + (2
\zeta r)^2}}.
\]
The transmissibility crosses unity at $r = \sqrt{2}$ and approaches
zero as $r \to \infty$ (Figure~\ref{fig:vol03ch06:transmissibility}).
The principle of vibration isolation is
based on this: a mounted mass that is to be isolated from support
vibration must be mounted on springs and dampers whose combined
natural frequency $\omega_n$ is well below the lowest driving
frequency $\omega_{\text{drive}}$, so that $r > \sqrt{2}$ and TR <
1. The further above $\sqrt{2}$, the better the isolation. The
isolation is independent of the magnitude of the support motion; it
depends only on the frequency ratio.

The flip side is that an isolated mass that experiences a transient
disturbance (an impulse, a startup) is set into oscillation at its
own $\omega_n$. The transient response decays at the system's
damping rate. Designing an isolator with low $\omega_n$ for steady-
state isolation can therefore produce poor transient performance;
the trade-off is the standard design balance in vibration isolation.

\input{volumes/03-force/ch06-vibrations/figures/fig-transmissibility.tex}

The transmissibility curve also exposes the role of damping in an
isolator. Above $r = \sqrt2$, more damping worsens isolation: a
damped mount transmits more high-frequency motion than an undamped
one, because the damping term $2\zeta r$ in the numerator grows with
frequency. Below $r = \sqrt2$, and at resonance in particular, damping
helps by capping the transmissibility peak. A practical isolator
therefore carries just enough damping to survive the resonance it
must pass through at startup, and no more. The helper program
\path{code/transmissibility.py} sizes an isolator from a mass, a
disturbance frequency, and a target transmissibility, and tabulates
the resulting curve across the operating band.

\begin{principle}[Resonance is the hazard and the tool]
The amplification of a forced response at the natural frequency,
$1/(2\zeta)$, is both the engineering hazard (a small force at the
right frequency produces a destructive response) and the
engineering tool (a tuned system responds preferentially to one
frequency and rejects others). Every dynamic system in the book is
designed either to avoid resonance (suspension bridges, building
columns, machine shafts) or to exploit it (musical instruments,
filters, resonance-based sensors). The design choice rests on the
ratio of driving-force amplitude, structural damping, and the
relative cost of the two failure modes.
\end{principle}

\section{Multi-degree-of-freedom systems and modal analysis}
\pathtag{standard}

A system with $n$ generalised coordinates has $n$ equations of
motion. For a linear system with mass matrix $M$, damping matrix $C$,
and stiffness matrix $K$ (all $n \times n$, symmetric, positive-
definite in the standard case), the equations of motion are
\[
M\ddot{\vec{q}} + C\dot{\vec{q}} + K\vec{q} = \vec{F}(t),
\]
a coupled system in the displacement vector $\vec{q}(t)$.

The undamped free vibration ($\vec{F} = 0$, $C = 0$) admits solutions
of the form $\vec{q}(t) = \vec{\phi} \cos(\omega t - \phi_0)$, with
$\vec{\phi}$ a constant vector. Substituting,
\[
(K - \omega^2 M)\vec{\phi} = \vec{0}.
\]
For non-trivial $\vec{\phi}$, the determinant must vanish:
\[
\det(K - \omega^2 M) = 0.
\]
This is the \engterm{generalised eigenvalue problem}. It has $n$
solutions for $\omega^2$, the \engterm{natural angular frequencies}
$\omega_1^2 \leq \omega_2^2 \leq \cdots \leq \omega_n^2$, with
corresponding eigenvectors $\vec{\phi}_1, \vec{\phi}_2, \ldots,
\vec{\phi}_n$, the \engterm{mode shapes}. Each mode shape describes
the relative displacements of the $n$ generalised coordinates when
the system oscillates purely in that mode; the natural frequency
gives the rate of oscillation of that mode
(Figure~\ref{fig:vol03ch06:2dof-modes}).

\input{volumes/03-force/ch06-vibrations/figures/fig-2dof-modes.tex}

The mode shapes are orthogonal with respect to the mass and
stiffness matrices: $\vec{\phi}_i^T M \vec{\phi}_j = 0$ and
$\vec{\phi}_i^T K \vec{\phi}_j = 0$ for $i \neq j$. The orthogonality
lets the engineer decouple the system: defining \engterm{modal
coordinates} $\eta_k$ by $\vec{q}(t) = \sum_k \eta_k(t) \vec{\phi}_k$
and substituting, the equations of motion separate into $n$
independent SDOF equations,
\[
m_k \ddot{\eta}_k + c_k \dot{\eta}_k + k_k \eta_k = \vec{\phi}_k^T
\vec{F}(t),
\]
with $m_k = \vec{\phi}_k^T M \vec{\phi}_k$, $k_k = \vec{\phi}_k^T K
\vec{\phi}_k$, and $\omega_k^2 = k_k/m_k$. The decoupling is exact for
the undamped case and approximate for damping that satisfies the
classical-damping condition (commonly the Rayleigh model $C = \alpha
M + \beta K$ for scalars $\alpha, \beta$).

Modal analysis is the working procedure for any structure with more
than two degrees of freedom: solve the eigenvalue problem once,
extract the natural frequencies and mode shapes, transform the
forcing to modal coordinates, solve $n$ SDOF problems, and transform
the response back to physical coordinates. The procedure is the basis
of structural dynamics, earthquake engineering, machinery vibration
analysis, and the modal-superposition method in finite-element
software.

\engterm{Participation factors}. Not every mode contributes equally
to the response to a given excitation. For a base excitation (the
support moves) on a many-degree-of-freedom system, each mode's
contribution is weighted by its participation factor $\Gamma_k =
\vec{\phi}_k^T M \vec{r} / m_k$, where $\vec{r}$ is the rigid-body
unit vector in the direction of the base motion. Modes with small
participation factors contribute little; in seismic analysis it is
standard to retain only enough modes to capture $90\,\%$ of the
total participating mass and to ignore the rest.

The program \path{code/modal_2dof.py} solves the generalised
eigenvalue problem for a given $M$ and $K$, returns the natural
frequencies and mass-normalised mode shapes, and verifies the
mass-orthogonality $\vec{\phi}_i^T M \vec{\phi}_j = \delta_{ij}$. Run
on the two-degree-of-freedom system of Calculation~\ref{ex:vol03ch06:2dof}
below, it returns mode shapes $[1, 2]^T$ at the lower frequency and
$[1, -1]^T$ at the higher, the in-phase and out-of-phase patterns of
Figure~\ref{fig:vol03ch06:2dof-modes}.

\begin{archetype}[Stability under a coincidence of frequencies]
The vibration engineer's recurring task is to keep a structure's
natural frequencies clear of the frequencies that drive it. A rotor
must not spin at a shaft critical speed; a floor must not resonate
with footfall; a circuit board's lowest mode must clear the vibration
spectrum of the vehicle it rides in. The archetype is a stability
question posed in the frequency domain: enumerate the natural
frequencies, enumerate the excitation frequencies, and verify that no
lightly damped mode coincides with a strong excitation inside the
operating envelope. When a coincidence cannot be avoided, the fallback
is to add damping or to detune one of the two frequencies by changing
mass or stiffness. The same enumeration recurs whether the system has
one degree of freedom, a handful, or a continuum.
\end{archetype}

\begin{mastery}
\textbf{The eigenvalue problem as the spectral structure of the
operator.} The generalised eigenvalue problem $K\vec{\phi} = \omega^2
M\vec{\phi}$ is the finite-dimensional shadow of a deeper statement:
the free vibration of any linear elastic system is governed by a
self-adjoint operator whose spectrum is the set of squared natural
frequencies and whose eigenfunctions are the mode shapes. For the
discrete system the operator is the matrix pair $(M, K)$; for the
continuous string it is the operator $-(T/\rho)\,d^2/dx^2$ acting on
functions that vanish at the ends; for the beam it is
$(EI/\rho A)\,d^4/dx^4$. In every case the self-adjointness (which
follows from the symmetry of $M$ and $K$, or from integration by
parts with the right boundary conditions) guarantees real
eigenvalues, orthogonal eigenfunctions, and a complete basis in which
any response expands. The Rayleigh quotient
\[
R(\vec{v}) = \frac{\vec{v}^T K \vec{v}}{\vec{v}^T M \vec{v}}
\]
bounds the lowest natural frequency from above for any trial vector
$\vec{v}$, and equals $\omega_1^2$ exactly when $\vec{v} =
\vec{\phi}_1$. This is the basis of the Rayleigh-Ritz method, which
estimates the fundamental frequency of a complicated structure from a
single plausible guess at its first mode shape, often to within a few
percent without solving the full eigenvalue problem.
\end{mastery}

\section{Continuous systems}\pathtag{mastery}

A vibrating string, beam, plate, or shell has infinitely many
degrees of freedom, indexed by position along (or across) the body.
The equations of motion are partial differential equations, and the
modal analysis generalises to a continuous-spectrum problem with
infinitely many natural frequencies and mode shapes.

\subsection{The vibrating string}

A taut string of length $L$, mass per unit length $\rho$, and
tension $T$ has the wave equation
\[
\frac{\partial^2 y}{\partial t^2} = c^2 \frac{\partial^2 y}{\partial
x^2}, \quad c = \sqrt{T/\rho},
\]
with $y(x, t)$ the lateral displacement and $c$ the wave speed
along the string. For a string fixed at both ends, the natural
modes are
\[
y_n(x, t) = A_n \sin\left(\frac{n\pi x}{L}\right) \cos(\omega_n t -
\phi_n),
\]
with natural frequencies
\[
\omega_n = \frac{n\pi c}{L} = \frac{n\pi}{L}\sqrt{T/\rho}, \quad n =
1, 2, 3, \ldots.
\]
The frequencies are an integer harmonic series; the fundamental is
$\omega_1 = \pi c/L$, the next harmonics are at $2\omega_1, 3\omega_1,
\ldots$. The integer harmonic structure is what makes vibrating
strings musically pleasing, the subject of chapter 7 of this volume.
The first three mode shapes appear in
Figure~\ref{fig:vol03ch06:mode-shapes}.

\input{volumes/03-force/ch06-vibrations/figures/fig-mode-shapes.tex}

\subsection{The Euler-Bernoulli beam}

A slender beam of length $L$, mass per unit length $\rho A$ (with
$\rho$ the material density and $A$ the cross-sectional area),
flexural rigidity $EI$, has the equation of motion for small
transverse vibrations
\[
\rho A \frac{\partial^2 y}{\partial t^2} + EI \frac{\partial^4 y}{
\partial x^4} = w(x, t),
\]
with $w(x, t)$ the distributed transverse force per unit length and
$y(x, t)$ the transverse deflection. The fourth-spatial-derivative
makes the beam dispersive: the wave speed depends on wavelength.
The boundary conditions at the beam's ends fix the natural-frequency
spectrum.

For a simply supported beam of length $L$, the natural frequencies
are
\[
\omega_n = (n\pi)^2 \sqrt{\frac{EI}{\rho A L^4}}, \quad n = 1, 2, 3,
\ldots.
\]
The frequency ratio is $1 : 4 : 9 : 16 : \cdots$, not harmonic, which
is why a struck beam sounds like a clang rather than a pitch.

For a cantilever beam, the natural-frequency-determining equation
is transcendental ($\cos\beta_n L \cosh\beta_n L + 1 = 0$, with
$\omega_n = \beta_n^2 \sqrt{EI/\rho A}$); the first few roots give
$\beta_1 L \approx 1.875$, $\beta_2 L \approx 4.694$, $\beta_3 L
\approx 7.855$, $\beta_4 L \approx 10.996$. The frequencies are not
in any simple ratio.

\subsection{Plates, shells, and three-dimensional bodies}

Two-dimensional and three-dimensional continuous systems have
correspondingly richer modal spectra. The mode shapes of a
rectangular plate, simply supported on all sides, are products of
sine functions in the two directions, with natural frequencies
proportional to $(m/L_x)^2 + (n/L_y)^2$ for integer mode numbers
$m, n$. The mode shapes of a circular drum head are Bessel functions
times sines or cosines. The mode shapes of a wine glass include
ring modes that are the basis of the "singing wineglass" experiment.
Chapter 9 of this volume develops plate and shell vibration in
more detail.

\begin{mastery}
\textbf{The cantilever-beam eigenproblem in full, and why its
frequencies are not harmonic.} The simply-supported beam gave a clean
sine basis because both the displacement and the bending moment vanish
at a pinned end, and $\sin(n\pi x/L)$ satisfies both. A cantilever
forces the issue: the clamped end fixes displacement and slope to zero,
$y = y' = 0$; the free end fixes shear and moment to zero, $y'' = y''' =
0$. The spatial equation from separating variables on the
Euler-Bernoulli beam is $\Phi'''' = \beta^4\Phi$ with $\beta^4 =
\omega^2\rho A/(EI)$, whose general solution mixes all four characters,
\[
\Phi(x) = a\cos\beta x + b\sin\beta x + c\cosh\beta x + d\sinh\beta x.
\]
The two clamped conditions kill the symmetric pairing and leave $\Phi =
a(\cos\beta x - \cosh\beta x) + b(\sin\beta x - \sinh\beta x)$. The two
free conditions then give a $2\times 2$ homogeneous system whose
determinant must vanish, which reduces to the transcendental
characteristic equation $\cos\beta L\,\cosh\beta L = -1$. Its roots
$\beta_n L = 1.875, 4.694, 7.855, 10.996,\ldots$ approach $(n -
\tfrac12)\pi$ for large $n$, so the natural frequencies $\omega_n =
(\beta_n L)^2\sqrt{EI/(\rho A L^4)}$ scale as the square of those roots:
$1 : 6.27 : 17.5 : 34.4 : \cdots$. The hyperbolic terms are the source
of the inharmonicity. They decay away from the clamped end and bend the
low roots away from the $(n - \tfrac12)\pi$ spacing that the pure
sinusoids would have given, which is why a struck cantilever rings with
a metallic clang rather than a pitched note. The mode shapes that go
with these roots are the products $\Phi_n(x)$ above evaluated at each
$\beta_n$; they are orthogonal under the mass weighting $\int_0^L \rho A
\,\Phi_m\Phi_n\,dx = 0$ for $m\neq n$, the continuous version of the
mass-orthogonality that decoupled the discrete system, and they form
the basis in which any cantilever response expands. The Rayleigh-Ritz
estimate of the previous mastery box uses exactly the lowest of these
as its trial function and recovers $1.875^2 = 3.516$ to the figures
quoted.
\end{mastery}

\section{Random vibration}\pathtag{mastery}

Real loadings are often not sinusoidal but random: turbulent wind on
a building, road roughness on a vehicle, ocean waves on an offshore
platform, jet-engine pressure on an aircraft skin. The driving
function $F(t)$ is then a random process, and the system response
is itself a random process. The analysis tool is the
\engterm{frequency response function} $H(\omega)$ of the system
combined with the \engterm{power spectral density} $S_F(\omega)$ of
the forcing.

For a linear SDOF system, the steady-state response to a stationary
random input has power spectral density
\[
S_x(\omega) = |H(\omega)|^2 S_F(\omega),
\]
where $|H(\omega)|^2 = 1/[k^2 ((1-r^2)^2 + (2\zeta r)^2)]$ is the
squared magnitude of the system's transfer function. The mean-square
response is the integral
\[
\langle x^2 \rangle = \int_{-\infty}^{\infty} S_x(\omega) \,d\omega.
\]
For white-noise input ($S_F(\omega) = S_0$ constant) and small
damping, the integral is dominated by the resonance region and
evaluates to
\[
\langle x^2 \rangle \approx \frac{\pi S_0}{2 k^2 \zeta \omega_n} =
\frac{\pi S_0}{2 c k}.
\]
The mean-square response is inversely proportional to the damping
$c$. For Gaussian forcing, the response is Gaussian, and the
probability of exceeding any threshold can be computed from
$\langle x^2 \rangle$ and the threshold's level.

\begin{mastery}
\textbf{Spectral moments and the rate of peaks.} The mean-square
response $\langle x^2 \rangle$ is the zeroth spectral moment $m_0 =
\int S_x(\omega)\,d\omega$. Higher moments carry more of the story.
The second moment $m_2 = \int \omega^2 S_x(\omega)\,d\omega$ is the
mean-square velocity, and the ratio $\sqrt{m_2/m_0}$ is the response's
characteristic angular frequency. For a stationary Gaussian process
with a narrow-band spectrum, the expected rate of upward crossings of
a level $a$ is
\[
\nu(a) = \frac{1}{2\pi}\sqrt{\frac{m_2}{m_0}}\,
\exp\!\left(-\frac{a^2}{2 m_0}\right),
\]
the Rice formula. The expected rate of peaks above $a$ follows, and
from it the expected fatigue damage per unit time once a stress-life
curve is supplied. This is the chain that turns a power spectral
density of road roughness or aerodynamic buffet into a predicted
fatigue life: spectrum $\to$ response spectrum via $|H(\omega)|^2$
$\to$ spectral moments $\to$ peak distribution $\to$ rainflow-equivalent
damage. Volume X develops the fatigue side; the vibration side is the
two-step $S_x = |H|^2 S_F$ and the spectral-moment integrals.
\end{mastery}

Random vibration analysis is the working tool of aerospace structural
design (jet-engine and aerodynamic-buffet response), automotive
suspension design (road-roughness response), offshore platform
design (ocean-wave response), and any engineering domain where the
load is irreducibly stochastic. The chapter touches the framework
only at introduction level; the full treatment is the subject of
chapter 14 of Volume X (maintenance failure under random loading)
and chapter 8 of Volume IX (system response to random inputs).

\section{Vibration measurement and signature analysis}\pathtag{standard}

A measurement of a vibrating system's response yields a time history
$x(t)$ from which the engineer can extract the system's modal
properties (natural frequencies, mode shapes, damping ratios) and
identify the operating regime. The standard tools:

\engterm{Accelerometers} are the most common sensors: a small
seismic mass on a piezoelectric or capacitive transducer, producing
a voltage proportional to acceleration over a defined bandwidth.
Piezoelectric accelerometers have flat response from a few Hz to
tens of kHz; capacitive accelerometers measure down to DC and are
the choice for low-frequency structural vibration (typically below
$50\,\si{\hertz}$). Range and resolution vary widely with the
specific sensor; modern industrial accelerometers report from
$\sim 10^{-6}\,g$ to $\sim 10^4\,g$ depending on model.

\engterm{Spectral analysis} converts the time history to its
frequency content. The fast Fourier transform (FFT) of $x(t)$ gives
the spectrum $X(\omega)$; the magnitude $|X(\omega)|^2$ is the
power spectrum. Peaks in the power spectrum identify natural
frequencies; the width of each peak (the "Q" of the resonance)
identifies damping. The frequency resolution of an FFT is the
inverse of the sample-record length; a $1\,\si{\second}$ record gives
$1\,\si{\hertz}$ resolution, a $10\,\si{\second}$ record $0.1\,\si{
\hertz}$. The companion program \path{code/fft_modal.py} reads a
recorded accelerometer time history, applies a Hann window to control
spectral leakage, computes the single-sided amplitude spectrum, finds
the dominant peaks, and estimates the damping ratio of the strongest
peak by the half-power bandwidth method. Run on the synthetic
cantilever decay in \path{data/cantilever_decay.csv} (a metre-stick
cantilever with a fundamental near $48\,\si{\hertz}$ and
$\zeta = 0.02$), it recovers the fundamental at $48\,\si{\hertz}$ and
a half-power damping ratio of $0.021$, within the frequency resolution
of the record.

\engterm{Modal identification} extracts mode shapes from
simultaneous measurements at many points. Excitation by impact
hammer (an instrumented hammer that produces a broadband impulse) or
by shaker (a controlled force at a chosen frequency or chirp)
produces a frequency response function $H(\omega)$ at each
measurement point. The peaks of $|H(\omega)|$ at each point identify
the natural frequencies; the relative amplitudes at the peak
identify the mode shape. Commercial modal-analysis software
performs this routinely on hundreds of measurement points.

\engterm{Signature analysis} compares a vibration spectrum to a
baseline (the same machine when it was new, the same structure
when it was first commissioned) and flags deviations
(Figure~\ref{fig:vol03ch06:spectrum}). Standard
machinery faults have standard signatures:

\input{volumes/03-force/ch06-vibrations/figures/fig-spectrum.tex}

\begin{itemize}[leftmargin=*]
  \item \engterm{Imbalance}: a peak at the rotation frequency $1
    \times$, possibly with overtones at $2\times$, $3\times$.
  \item \engterm{Misalignment}: peaks at $1\times$ and $2\times$
    with $2\times$ comparable to or larger than $1\times$.
  \item \engterm{Bearing wear}: broad-band noise increase, plus
    specific peaks at the bearing's race-defect frequencies
    (computed from the bearing's geometry).
  \item \engterm{Looseness}: subharmonic peaks at $0.5\times$ or
    impulsive responses with multiple harmonics.
  \item \engterm{Resonance}: a peak that doesn't move with rotation
    speed and is sharp (small bandwidth).
\end{itemize}

The bearing-band peaks are not arbitrary. A rolling-element bearing
generates four characteristic defect frequencies, each a non-integer
multiple of the shaft speed $f_s$ fixed by the bearing geometry: the
ball-pass frequency of the outer race, $\text{BPFO} = (N/2) f_s (1 -
(d/D)\cos\alpha)$; the ball-pass frequency of the inner race,
$\text{BPFI} = (N/2) f_s (1 + (d/D)\cos\alpha)$; the ball-spin
frequency; and the fundamental train (cage) frequency. Here $N$ is the
number of rolling elements, $d$ the element diameter, $D$ the pitch
diameter, and $\alpha$ the contact angle. A defect on the outer race
strikes at BPFO, on the inner race at BPFI, and so on, so a line at a
known geometric frequency localises the fault to a specific surface of
a specific bearing before the machine is opened. The non-integer ratios
($\text{BPFO}/f_s$ is typically near $3$ to $5$ for common bearings,
and never an integer) are what let signature analysis separate a
bearing line from the running-speed orders that surround it.

The signatures let the maintenance engineer diagnose the fault from
a vibration measurement without disassembling the machine. The
discipline of routine vibration monitoring, with computer-stored
spectra trended over months and years, is now standard practice
for critical industrial machinery and is the basis of
condition-based maintenance. A representative machine spectrum, with
labelled imbalance, misalignment, looseness, and bearing-band peaks at
a running speed of $25\,\si{\hertz}$ ($1500\,\text{rpm}$), is provided
in \path{data/machine_spectrum.csv} for the reader to load and
classify.

\begin{estimation}
\textbf{Estimate the natural frequency of a $300\,\si{\milli\meter}$
metre stick clamped at one end as a cantilever, made of pine wood
of cross-section $25\,\si{\milli\meter} \times 6\,\si{\milli\meter}$.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
For a cantilever of length $L$, mass per unit length $\rho A$, and
flexural rigidity $EI$, the first natural angular frequency is
\[
\omega_1 = (1.875)^2 \sqrt{\frac{EI}{\rho A L^4}}.
\]
For pine, $E \approx 10\,\text{GPa}$, density $\rho \approx 500\,\si{
\kilogram\per\meter\cubed}$. The cross-section is $b \times h = 25
\times 6\,\si{\milli\meter}$, so $A = 1.5 \times 10^{-4}\,\si{\meter
\squared}$, and assuming the load is in the strong direction
(perpendicular to the $25\,\si{\milli\meter}$ width), $I = bh^3/12 =
0.025 \times (0.006)^3/12 = 4.5 \times 10^{-10}\,\si{\meter\tothe{4}}$.
Then
\[
EI = 10^{10} \times 4.5 \times 10^{-10} = 4.5\,\si{\newton\meter\squared},
\]
\[
\rho A = 500 \times 1.5 \times 10^{-4} = 0.075\,\si{\kilogram\per\meter},
\]
\[
\frac{EI}{\rho A L^4} = \frac{4.5}{0.075 \times (0.3)^4} = \frac{4.5}{
6.07 \times 10^{-4}} \approx 7{,}400\,\si{\per\second\squared}.
\]
So $\omega_1 \approx (3.516) \sqrt{7{,}400} \approx 3.516 \times 86
\approx 303\,\si{\radian\per\second}$, and $f_1 = \omega_1/(2\pi)
\approx 48\,\si{\hertz}$.

Estimate: roughly $50\,\si{\hertz}$.

The postmortem: the answer is sensitive to the orientation of the
cross-section (strong or weak direction), to the elastic modulus
(pine varies from $8$ to $14\,\text{GPa}$ depending on species and
grain direction), and to the exact length. A measurement on a real
metre stick clamped in a vice typically returns a fundamental in
the range $30$ to $60\,\si{\hertz}$, in agreement. The project of
this chapter is to measure this for the reader's own apparatus and
compare to the prediction.
\end{estimation}

\section{Worked examples}\pathtag{standard}

The formulas of the preceding sections become working tools only when
they are run end to end on numbers. The examples below carry a
calculation from the physical statement through the algebra to a
checked result, in the order a practising engineer would take them.

\subsection{Free decay: extracting damping from a ring-down}

A steel cantilever is struck and its tip acceleration recorded. Two
displacement peaks twelve cycles apart measure $4.20\,\si{\milli
\meter}$ and $1.05\,\si{\milli\meter}$, and the time between the first
and last of the twelve cycles is $0.240\,\si{\second}$. Find the
damped natural frequency, the damping ratio, and the undamped natural
frequency.

\textbf{Damped frequency.} Twelve cycles in $0.240\,\si{\second}$
gives a damped period $T_d = 0.240/12 = 0.0200\,\si{\second}$, so
$f_d = 50.0\,\si{\hertz}$ and $\omega_d = 2\pi f_d = 314\,\si{\radian
\per\second}$.

\textbf{Logarithmic decrement.} The amplitude ratio over twelve
cycles is $4.20/1.05 = 4.00$, so the per-cycle decrement is
\[
\delta = \frac{1}{12}\ln(4.00) = \frac{1.386}{12} = 0.1155.
\]

\textbf{Damping ratio.} Inverting $\delta = 2\pi\zeta/\sqrt{1-\zeta^2}$,
\[
\zeta = \frac{\delta}{\sqrt{4\pi^2 + \delta^2}} = \frac{0.1155}{\sqrt{
39.48 + 0.0133}} = \frac{0.1155}{6.284} = 0.01838.
\]
The light-damping approximation $\zeta \approx \delta/(2\pi) =
0.01838$ agrees to four figures, as expected for $\zeta$ well below
$0.1$.

\textbf{Undamped frequency.} Since $\omega_d = \omega_n\sqrt{1 -
\zeta^2}$,
\[
\omega_n = \frac{\omega_d}{\sqrt{1 - \zeta^2}} = \frac{314}{\sqrt{1 -
0.000338}} = 314.05\,\si{\radian\per\second}.
\]
The damping shifts the frequency by $0.017\,\%$, far below the
measurement's resolution. For light damping the measured $f_d$ is the
natural frequency to every practical purpose.

\subsection{Forced response away from and at resonance}

A SDOF system has $m = 8\,\si{\kilogram}$, $k = 3200\,\si{\newton\per
\meter}$, and $\zeta = 0.04$. It is driven by a harmonic force of
amplitude $F_0 = 20\,\text{N}$. Find the steady-state amplitude at
$\omega = 0.5\,\omega_n$, at $\omega = \omega_n$, and at $\omega =
2\,\omega_n$.

\textbf{Baseline quantities.} The natural frequency is $\omega_n =
\sqrt{k/m} = \sqrt{3200/8} = 20.0\,\si{\radian\per\second}$. The
static deflection is $F_0/k = 20/3200 = 6.25\,\si{\milli\meter}$.

\textbf{Half-resonance, $r = 0.5$.}
\[
X = \frac{F_0/k}{\sqrt{(1 - 0.25)^2 + (2\cdot 0.04\cdot 0.5)^2}} =
\frac{6.25}{\sqrt{0.5625 + 0.0016}} = \frac{6.25}{0.7510} = 8.32\,\si{
\milli\meter}.
\]
The amplification is $1.33$: below resonance the response modestly
exceeds the static deflection.

\textbf{Resonance, $r = 1$.}
\[
X = \frac{F_0/k}{2\zeta} = \frac{6.25}{0.08} = 78.1\,\si{\milli\meter}.
\]
The amplification is $1/(2\zeta) = 12.5$, an order of magnitude over
the static deflection from the same force.

\textbf{Twice resonance, $r = 2$.}
\[
X = \frac{6.25}{\sqrt{(1 - 4)^2 + (2\cdot 0.04\cdot 2)^2}} =
\frac{6.25}{\sqrt{9 + 0.0256}} = \frac{6.25}{3.004} = 2.08\,\si{\milli
\meter}.
\]
The amplification is $0.33$, close to the high-frequency asymptote
$1/r^2 = 0.25$: above resonance the mass cannot follow the force. The
same force produces responses spanning a factor of $38$ across these
three frequencies, the entire point of resonance analysis.

\subsection{Two-degree-of-freedom modes by hand}

Find the natural frequencies and mode shapes of a system with
$M = \operatorname{diag}(2, 1)\,\si{\kilogram}$ and stiffness matrix
\[
K = \begin{pmatrix} 3 & -1 \\ -1 & 1 \end{pmatrix}\,\si{\newton\per
\meter}.
\]

\textbf{Characteristic equation.} Set $\det(K - \omega^2 M) = 0$:
\[
\det\begin{pmatrix} 3 - 2\omega^2 & -1 \\ -1 & 1 - \omega^2
\end{pmatrix} = (3 - 2\omega^2)(1 - \omega^2) - 1 = 0.
\]
Expanding, $2\omega^4 - 5\omega^2 + 3 - 1 = 2\omega^4 - 5\omega^2 + 2
= 0$. With $u = \omega^2$,
\[
2u^2 - 5u + 2 = 0 \implies u = \frac{5 \pm \sqrt{25 - 16}}{4} =
\frac{5 \pm 3}{4} = 2 \text{ or } 0.5.
\]

\textbf{Natural frequencies.} $\omega_1 = \sqrt{0.5} = 0.707\,\si{
\radian\per\second}$ and $\omega_2 = \sqrt{2} = 1.414\,\si{\radian\per
\second}$.

\textbf{Mode shapes.} For each $\omega^2$, solve $(K - \omega^2 M)
\vec{\phi} = \vec{0}$. The first row gives $(3 - 2\omega^2)\phi_1 =
\phi_2$.

For $\omega_1^2 = 0.5$: $(3 - 1)\phi_1 = \phi_2$, so $\phi_2 =
2\phi_1$ and $\vec{\phi}_1 = [1, 2]^T$. Both masses move the same
direction, the lighter one further: the in-phase mode.

For $\omega_2^2 = 2$: $(3 - 4)\phi_1 = \phi_2$, so $\phi_2 = -\phi_1$
and $\vec{\phi}_2 = [1, -1]^T$. The masses move in opposite
directions: the out-of-phase mode. These reproduce the numerical
output of \path{code/modal_2dof.py} and the schematic of
Figure~\ref{fig:vol03ch06:2dof-modes}.

\textbf{Orthogonality check.} $\vec{\phi}_1^T M \vec{\phi}_2 = [1, 2]
\operatorname{diag}(2,1) [1, -1]^T = (1)(2)(1) + (2)(1)(-1) = 2 - 2 =
0$, confirming mass-orthogonality.

\subsection{Sizing a vibration isolator}

A $100\,\si{\kilogram}$ machine runs at $50\,\si{\hertz}$ and must
transmit no more than $5\,\%$ of its disturbance force to the floor.
Specify the isolator natural frequency, the total spring stiffness,
and the damping.

\textbf{Required frequency ratio.} For light damping the
transmissibility above resonance is approximately $\mathrm{TR}
\approx 1/(r^2 - 1)$. Setting this to $0.05$,
\[
r^2 - 1 = \frac{1}{0.05} = 20 \implies r = \sqrt{21} = 4.58.
\]

\textbf{Isolator natural frequency.} With the disturbance at $50\,
\si{\hertz}$,
\[
f_n = \frac{f_{\text{drive}}}{r} = \frac{50}{4.58} = 10.9\,\si{\hertz}.
\]

\textbf{Spring stiffness.} $\omega_n = 2\pi f_n = 68.6\,\si{\radian
\per\second}$, so
\[
k = m\omega_n^2 = 100 \times (68.6)^2 = 4.70 \times 10^5\,\si{\newton
\per\meter},
\]
distributed across the chosen number of mounts (four mounts at $118\,
\si{\kilo\newton\per\meter}$ each).

\textbf{Damping and the startup penalty.} A small damping ratio,
$\zeta \approx 0.05$, keeps the resonant peak at $1/(2\zeta) = 10$ as
the machine spins up through $10.9\,\si{\hertz}$ while preserving
high-frequency isolation. The exact transmissibility at $r = 4.58$,
$\zeta = 0.05$ is $0.055$, confirming the $5\,\%$ target. These
numbers match \path{code/transmissibility.py}.

\subsection{Base excitation: a chassis-mounted instrument}

An instrument of mass $m = 3\,\si{\kilogram}$ is mounted on a vehicle
chassis through elastomer feet of combined stiffness $k = 1.2\times
10^4\,\si{\newton\per\meter}$ and damping ratio $\zeta = 0.08$. The
chassis vibrates vertically with displacement amplitude $Y = 2\,\si{
\milli\meter}$ at the road-input frequency $18\,\si{\hertz}$. Find the
instrument's absolute motion amplitude and the relative motion across
the feet, the quantity that loads the elastomer.

\textbf{Mount natural frequency.} $\omega_n = \sqrt{k/m} = \sqrt{1.2
\times 10^4/3} = 63.2\,\si{\radian\per\second}$, so $f_n = 10.1\,\si{
\hertz}$. The drive at $18\,\si{\hertz}$ gives $r = 18/10.1 = 1.78$,
above $\sqrt2$, so the mount isolates rather than amplifies.

\textbf{Absolute motion (transmissibility).} The base-excitation
transmissibility is the same TR$(\omega)$ used for force isolation:
\[
\frac{X}{Y} = \sqrt{\frac{1 + (2\zeta r)^2}{(1 - r^2)^2 + (2\zeta
r)^2}} = \sqrt{\frac{1 + (2\cdot 0.08\cdot 1.78)^2}{(1 - 3.17)^2 +
(0.285)^2}} = \sqrt{\frac{1.081}{4.79}} = 0.475.
\]
The instrument moves with amplitude $X = 0.475 \times 2 = 0.95\,\si{
\milli\meter}$, less than half the chassis motion.

\textbf{Relative motion (what the feet feel).} The relative
displacement $z = x - y$ across the mount has its own transfer
function,
\[
\frac{Z}{Y} = \frac{r^2}{\sqrt{(1 - r^2)^2 + (2\zeta r)^2}} =
\frac{3.17}{\sqrt{4.79}} = 1.45,
\]
so $Z = 1.45 \times 2 = 2.90\,\si{\milli\meter}$. The relative
deflection exceeds the base motion because operation is above
resonance: the instrument lags the chassis, and the feet take up the
difference. The two readings answer two different design questions. The
absolute motion $X$ sets whether the instrument functions; the relative
motion $Z$ sets the fatigue duty on the elastomer and the clearance the
mount must allow. A design that isolates the instrument well (small $X$)
can still tear its mounts (large $Z$), the trap in any above-resonance
isolation scheme.

\subsection{Rotating unbalance in a benchtop centrifuge}

A benchtop centrifuge rotor of total mounted mass $m = 6\,\si{
\kilogram}$ carries a residual unbalance of $m_e e = 8\,\si{\gram}
\cdot\si{\milli\meter} = 8\times 10^{-6}\,\si{\kilogram}\cdot\si{
\meter}$ (an $8\,\si{\gram}$ offset at $1\,\si{\milli\meter}$, or any
equal product). The machine sits on isolators giving $f_n = 12\,\si{
\hertz}$ and $\zeta = 0.10$. Find the whirl amplitude and the force
into the bench at the operating speed of $6000\,\text{rpm}$.

\textbf{Operating speed and ratio.} $6000\,\text{rpm} = 100\,\si{
\hertz}$, so $\omega = 2\pi\cdot 100 = 628\,\si{\radian\per\second}$
and $r = 100/12 = 8.33$, far above resonance.

\textbf{Whirl amplitude.} For rotating-unbalance excitation the
displacement amplitude is
\[
X = \frac{m_e e}{m}\cdot\frac{r^2}{\sqrt{(1 - r^2)^2 + (2\zeta r)^2}}.
\]
With $r = 8.33$, $r^2 = 69.4$, and $(1 - r^2)^2 = (68.4)^2 = 4{,}681$,
$(2\zeta r)^2 = (1.67)^2 = 2.78$, the dynamic factor is $69.4/\sqrt{
4{,}684} = 69.4/68.4 = 1.014$, close to the high-speed limit of unity
in Figure~\ref{fig:vol03ch06:rotating-unbalance}. Then
\[
X = \frac{8\times 10^{-6}}{6}\times 1.014 = 1.35\times 10^{-6}\,\si{
\meter} = 1.35\,\si{\micro\meter}.
\]
The rotor orbits its own mass centre with a radius near the bare
eccentricity scaled by the mass ratio, the signature of supercritical
running.

\input{volumes/03-force/ch06-vibrations/figures/fig-rotating-unbalance.tex}

\textbf{Transmitted force.} The force delivered to the bench through
the isolators is the transmissibility times the rotating-force
amplitude $F_0 = m_e e\,\omega^2$. The unbalance force is $F_0 = 8
\times 10^{-6}\times (628)^2 = 3.15\,\text{N}$. Above resonance the
isolators pass a small fraction; with TR $= \sqrt{(1 + (2\zeta
r)^2)/((1-r^2)^2 + (2\zeta r)^2)} = \sqrt{(1 + 2.78)/4{,}684} =
0.0284$, the transmitted force is $0.0284\times 3.15 = 0.089\,\text{N}$.
The lesson sits in the speed dependence: the unbalance force grows as
$\omega^2$, so doubling the speed quadruples it, and the isolator
earns its place by holding the transmitted fraction near three percent
across the running band. The same $m_e e$ at a sub-resonant speed near
$12\,\si{\hertz}$ would resonate and transmit far more, the reason a
spin-up that lingers at the isolator frequency is the dangerous part of
the duty cycle.

\subsection{A tuned absorber that kills a resonance}

A grinding head behaves as a SDOF system with $m_1 = 20\,\si{
\kilogram}$ and $k_1 = 3.16\times 10^5\,\si{\newton\per\meter}$, so its
natural frequency is $\omega_1 = \sqrt{k_1/m_1} = 125.7\,\si{\radian
\per\second}$ ($f_1 = 20.0\,\si{\hertz}$). In service it is driven at
exactly $20\,\si{\hertz}$ by a process force and rings badly. Design a
tuned absorber that holds the head still at $20\,\si{\hertz}$, and
state the price paid.

\textbf{Undamped absorber tuning.} A dynamic vibration absorber adds an
auxiliary mass $m_2$ on a spring $k_2$ tuned so that $\omega_2 =
\sqrt{k_2/m_2} = \omega_{\text{drive}}$. At that frequency the absorber
mass oscillates exactly out of phase with the applied force and its
spring force cancels the drive, holding the primary mass at zero. Pick a
mass ratio $\mu = m_2/m_1 = 0.05$, so $m_2 = 1\,\si{\kilogram}$. Tuning
$\omega_2 = \omega_{\text{drive}} = 125.7\,\si{\radian\per\second}$
fixes
\[
k_2 = m_2\,\omega_2^2 = 1\times (125.7)^2 = 1.58\times 10^4\,\si{
\newton\per\meter}.
\]

\textbf{The two new resonances.} Cancelling the response at one
frequency creates two new natural frequencies straddling it, the roots
of the two-degree-of-freedom characteristic equation. For equal tuning
($\omega_2 = \omega_1$) the two new frequency ratios are
\[
\left(\frac{\omega}{\omega_1}\right)^2 = 1 + \frac{\mu}{2} \pm
\sqrt{\mu + \frac{\mu^2}{4}},
\]
which for $\mu = 0.05$ gives $0.806$ and $1.244$, that is new
resonances at $f \approx 18.0\,\si{\hertz}$ and $22.3\,\si{\hertz}$.
The absorber turns one peak at $20\,\si{\hertz}$ into a deep notch at
$20\,\si{\hertz}$ flanked by two peaks at $18$ and $22\,\si{\hertz}$,
the split shown in Figure~\ref{fig:vol03ch06:absorber}. The design is
sound only if the drive stays near $20\,\si{\hertz}$; a drive that
wanders toward $18$ or $22\,\si{\hertz}$ finds a new resonance, which
is why the undamped absorber suits a fixed-frequency excitation and the
damped absorber (the tuned mass damper of the building exercise) suits a
band.

\input{volumes/03-force/ch06-vibrations/figures/fig-absorber.tex}

\textbf{The absorber's own amplitude.} At the operating frequency the
primary mass is still, so the entire applied force is reacted by the
absorber spring. The absorber displacement is $X_2 = -F_0/k_2$, larger
than the static deflection the primary would have shown by the factor
$k_1/k_2 = 3.16\times 10^5/1.58\times 10^4 = 20 = 1/\mu$. A light
absorber moves far; its stroke and spring stress set the practical
lower limit on $\mu$.

\subsection{Critical speed of a shaft from a Campbell diagram}

A pump shaft carries a rotor whose first lateral (whirl) bending mode
sits at $42\,\si{\hertz}$, measured at rest. The machine is variable
speed and must run somewhere between $3000$ and $3600\,\text{rpm}$.
Locate the critical speeds, decide whether the shaft runs subcritically
or supercritically with respect to its first mode, and state the margin.

\textbf{Running-speed orders.} The $1\times$ excitation rises with
speed as $f_{1\times} = N/60$ in hertz, and the $2\times$ at
$f_{2\times} = 2N/60$. A critical speed is a speed at which an order
crosses a natural frequency. The first mode at $42\,\si{\hertz}$ is hit
by the $1\times$ order when $N/60 = 42$, that is at
\[
N_{\text{crit},1\times} = 42\times 60 = 2520\,\text{rpm},
\]
and by the $2\times$ order when $2N/60 = 42$, at $N = 1260\,\text{rpm}$.

\textbf{Placement against the operating band.} Both crossings,
$1260$ and $2520\,\text{rpm}$, lie below the operating band of $3000$ to
$3600\,\text{rpm}$, so inside the band the running speed is above the
first whirl critical: the shaft runs supercritically with respect to
its first mode. The nearest critical, $2520\,\text{rpm}$, sits a factor
$3000/2520 = 1.19$ below the bottom of the band, a $19\,\%$ separation.
A common acceptance rule keeps the operating band at least $15\,\%$
clear of any critical, so the margin is adequate but not generous; a
soft-start that walks the speed through $2520\,\text{rpm}$ quickly keeps
the resonant dwell short. The Campbell diagram of
Figure~\ref{fig:vol03ch06:campbell} is the picture of this reasoning,
with the bending modes drawn flat and the orders rising as straight
lines from the origin.

\input{volumes/03-force/ch06-vibrations/figures/fig-campbell.tex}

\textbf{The supercritical advantage.} Above the first critical the rotor
spins about its own mass centre rather than its geometric centre, so the
self-balancing effect of supercritical running keeps the whirl orbit
small once the machine is through the critical. The price is the
transit: every start and stop drags the speed through $2520\,\text{rpm}$
and excites the lightly damped first mode in passing. The design
discipline is to clear the steady-state band of every critical and to
manage the transient crossings, not to demand a band with no critical
anywhere below it, which a high-speed machine cannot give.

\subsection{Seismic demand from a response spectrum}

A four-storey steel frame reduces, in its first sway mode, to an
equivalent SDOF system with first-mode period $T_1 = 2.0\,\si{\second}$,
a participating modal mass of $M_1 = 1.8\times 10^6\,\si{\kilogram}$
(about $80\,\%$ of the building mass), and modal damping
$\zeta = 0.05$. The site design response spectrum is the one plotted in
Figure~\ref{fig:vol03ch06:response-spectrum}. Find the first-mode base
shear without integrating a single ground-motion time history.

\input{volumes/03-force/ch06-vibrations/figures/fig-response-spectrum.tex}

\textbf{Read the spectrum.} The structure's period $T_1 = 2.0\,\si{
\second}$ falls on the falling velocity branch of the spectrum, where
$S_a = 0.45/T$ in units of $g$. At $T_1 = 2.0\,\si{\second}$,
\[
S_a = \frac{0.45}{2.0} = 0.225\,g = 0.225\times 9.81 = 2.21\,\si{\meter
\per\second\squared}.
\]
A stiff structure on the same site, with a period on the acceleration
plateau, would have drawn the full $0.9\,g$; the flexible frame filters
the high-frequency ground content and sees a quarter of that.

\textbf{First-mode base shear.} The peak base shear carried by the
first mode is the participating mass times the pseudo-acceleration,
\[
V_1 = M_1\,S_a = 1.8\times 10^6\times 2.21 = 3.97\times 10^6\,\text{N}
\approx 4.0\,\text{MN}.
\]
This is the force the lateral system must carry at the base from the
first mode alone. Higher modes add their own contributions, each read
from the same spectrum at its own period and combined by a modal-
combination rule (the square-root-of-sum-of-squares rule when the modal
periods are well separated). The response spectrum collapses the entire
dynamic analysis into a lookup per mode, which is why it is the
working tool of earthquake-resistant design and the reason a structural
engineer reasons about a building's fundamental period before anything
else.

\subsection{Damping and frequency from a measured frequency response
function}

A modal test on a bracket excites it with an instrumented hammer and
records the frequency response function $|H(\omega)|$ at the response
point. The magnitude curve shows a single isolated peak at
$f_n = 86.0\,\si{\hertz}$ with peak magnitude $H_{\text{peak}}$. The two
frequencies at which the magnitude falls to $H_{\text{peak}}/\sqrt2$ (the
half-power points) are measured at $84.3$ and $87.8\,\si{\hertz}$.
Extract the damping ratio and the modal stiffness, given a measured
static compliance $H(0) = 2.0\times 10^{-7}\,\si{\meter\per\newton}$.

\textbf{Half-power bandwidth.} For a lightly damped SDOF resonance the
fractional bandwidth between the two half-power points equals twice the
damping ratio,
\[
2\zeta = \frac{f_2 - f_1}{f_n} = \frac{87.8 - 84.3}{86.0} = \frac{3.5}{
86.0} = 0.0407,
\]
so $\zeta = 0.0203$, near two percent, the range of a bolted aluminium
bracket. The half-power method is the standard reading of damping from
a measured curve, and it needs only three frequencies off the plot, not
a fit to the whole resonance.

\textbf{Modal stiffness.} The low-frequency limit of the magnitude is
the static compliance $H(0) = 1/k$, so
\[
k = \frac{1}{H(0)} = \frac{1}{2.0\times 10^{-7}} = 5.0\times 10^6\,\si{
\newton\per\meter}.
\]

\textbf{Quality factor check.} The peak magnitude is the static
compliance amplified by the quality factor $Q = 1/(2\zeta) = 24.6$, so
the model predicts $H_{\text{peak}} = Q\,H(0) = 24.6\times 2.0\times
10^{-7} = 4.9\times 10^{-6}\,\si{\meter\per\newton}$. Reading the peak
height off the measured curve and comparing to this figure is the
internal consistency check that the resonance is a clean single mode and
not two merged peaks; a measured peak far below $Q\,H(0)$ is the warning
that the half-power points have been read across two overlapping modes.
The same \path{code/fft_modal.py} half-power routine that processed the
cantilever decay reads these three frequencies off an FRF and returns
the same $\zeta$.

\begin{mastery}
\textbf{State-space form and why it carries non-classical damping.} The
second-order modal decoupling of the previous section works only when
the damping matrix $C$ shares the eigenvectors of the undamped problem,
the classical-damping condition. Real structures with discrete dampers,
soil contact, or joined substructures violate it, and the modes no
longer decouple in the real domain. The state-space form handles the
general case. Stack displacement and velocity into a state vector
$\vec{z} = [\vec{q}^T, \dot{\vec{q}}^T]^T$ and write the second-order
system $M\ddot{\vec{q}} + C\dot{\vec{q}} + K\vec{q} = \vec{F}$ as the
first-order system
\[
\dot{\vec{z}} = A\vec{z} + B\vec{u}, \qquad
A = \begin{pmatrix} 0 & I \\ -M^{-1}K & -M^{-1}C \end{pmatrix}.
\]
The eigenvalues of the $2n\times 2n$ matrix $A$ come in complex
conjugate pairs $\lambda_k = -\zeta_k\omega_k \pm i\omega_k\sqrt{1 -
\zeta_k^2}$, from which the modal frequency and damping read off
directly as $\omega_k = |\lambda_k|$ and $\zeta_k = -\operatorname{Re}
(\lambda_k)/|\lambda_k|$. The eigenvectors are complex, which is the
mathematical signature of non-classical damping: the physical
coordinates no longer reach their peaks at the same instant, and a mode
shape now carries a phase as well as an amplitude. The state-space form
is also the bridge to control: $A$ is the open-loop dynamics matrix, $B$
the actuator map, and an active vibration controller designs a feedback
law $\vec{u} = -K_{\text{fb}}\vec{z}$ that places the closed-loop
eigenvalues where the passive structure could not reach them, the same
algebra that Volume IX develops for feedback control. Modal
superposition and state-space are the two standard routes through a
many-degree-of-freedom system; the first is cheaper and assumes
classical damping, the second is general and is the one finite-element
solvers fall back to when the damping will not diagonalise.
\end{mastery}

\begin{estimation}
\textbf{Estimate the first sway natural frequency of a forty-storey
office tower.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A working rule of thumb among structural engineers puts a building's
fundamental sway period at roughly $0.1\,\si{\second}$ per storey, so
a forty-storey tower has $T_1 \approx 4\,\si{\second}$ and $f_1
\approx 0.25\,\si{\hertz}$. The rule can be backed by a beam model.
Treat the tower as a uniform cantilever of height $H = 40 \times 3.5 =
140\,\si{\meter}$, bending stiffness $EI$, and mass per unit height
$\rho A$. The first cantilever frequency is $\omega_1 = (1.875)^2
\sqrt{EI/(\rho A H^4)}$. Engineering a plausible $EI$ from a steel
moment frame is involved, so the period rule is the faster route: a
modern tall building of this height typically reports $f_1$ between
$0.15$ and $0.3\,\si{\hertz}$. The estimate of $0.25\,\si{\hertz}$
sits squarely in that band.

\smallskip
The postmortem: the period rule encodes the empirical fact that
taller buildings are not proportionally stiffer, so the fundamental
period grows roughly linearly with height. The danger frequency is
the match between this $0.25\,\si{\hertz}$ sway mode and the dominant
energy in the wind-gust spectrum, which peaks below $0.1\,\si{\hertz}$
but carries energy across the band, and the match with vortex
shedding off the building's own faces. The tuned mass damper of the
chapter's design exercise targets exactly this mode.
\end{estimation}

\section{Case study: from a measured spectrum to an isolated
machine}\pathtag{standard}

The chapter's tools earn their keep when a single problem runs from a
measurement through a diagnosis to a design and ends with a check. We
follow one boiler-feed pump through that full arc: a recorded vibration
spectrum, a fault diagnosis read from the spectrum, a corrective action,
and an isolation mount sized to keep the residual disturbance out of the
foundation. The numbers carry through from the first reading to the last
verification.

\subsection{The machine and the complaint}

A horizontal boiler-feed pump runs at $1440\,\text{rpm}$, driven by a
four-pole induction motor on a shared baseplate. The maintenance crew
reports a low drumming that has grown over the past quarter and a hot
bearing at the coupling end. The pump rotor mass on its bearings is
$m = 180\,\si{\kilogram}$; the baseplate sits on a concrete plinth. The
running speed in hertz is
\[
f_{\text{run}} = \frac{1440}{60} = 24.0\,\si{\hertz},
\]
the $1\times$ order that anchors every reading that follows. A triaxial
accelerometer at the coupling-end bearing housing records a ten-second
velocity time history, sampled at $2\,\si{\kilo\hertz}$. The program
\path{code/fft_modal.py} windows the record, transforms it, and
returns the single-sided velocity spectrum of
Figure~\ref{fig:vol03ch06:casestudy-spectrum}.

\input{volumes/03-force/ch06-vibrations/figures/fig-casestudy-spectrum.tex}

\subsection{Reading the spectrum}

The spectrum carries four features, each a line in the fault table of
Section 6.7. We read them in order of running-speed order.

\textbf{The $1\times$ line at $24\,\si{\hertz}$} measures $2.6\,\si{
\milli\meter\per\second}$ RMS. A $1\times$ peak alone points at
imbalance, and on a pump that has run clean for years a slowly rising
$1\times$ would mean product buildup or erosion on the impeller.

\textbf{The $2\times$ line at $48\,\si{\hertz}$} measures $5.9\,\si{
\milli\meter\per\second}$ RMS, larger than the $1\times$. A $2\times$
that exceeds the $1\times$ is the signature of shaft misalignment, not
imbalance: the coupling, forced through two high and two low spots per
revolution by an angular or parallel offset between motor and pump
shafts, injects energy at twice the running speed. The hot
coupling-end bearing corroborates a misalignment, which loads that
bearing with a steady side force.

\textbf{The $3\times$ line at $72\,\si{\hertz}$} is small, $1.5\,\si{
\milli\meter\per\second}$, consistent with a moderate misalignment that
also seeds a third order.

\textbf{The broadband zone near $120$ to $160\,\si{\hertz}$} with a
discrete line at $146\,\si{\hertz}$ is not an integer order of the
running speed ($146/24 = 6.08$). A non-synchronous line riding on a
raised broadband floor is the early signature of a rolling-element
bearing defect, whose race-defect frequencies follow from the bearing
geometry and rarely land on an integer order. The misalignment side
load is the plausible root cause of the bearing's accelerated wear.

The diagnosis: a misaligned coupling driving a $2\times$ side force,
which both produces the drumming and overloads the coupling-end
bearing, whose early defect now shows in the bearing band. Imbalance is
present but secondary.

\subsection{The corrective action and what it leaves behind}

The first fix is mechanical, not dynamic. The crew shims and aligns the
motor to the pump within the coupling manufacturer's tolerance, which
collapses the $2\times$ line and removes the bearing side load; the
defective bearing is replaced. A field balance trims the residual
$1\times$. After the work, the $2\times$ falls below $1\,\si{\milli
\meter\per\second}$, the $1\times$ to $0.8\,\si{\milli\meter\per
\second}$, and the bearing band returns to the noise floor.

A residual $1\times$ unbalance force remains, as it always does. Take
the post-balance residual as $m_e e = 4\times 10^{-4}\,\si{\kilogram}
\cdot\si{\meter}$ (a balance grade in reach of field work on a pump
this size). The rotating force at running speed is
\[
F_0 = m_e e\,\omega^2 = 4\times 10^{-4}\times (2\pi\cdot 24)^2 = 4
\times 10^{-4}\times 2.27\times 10^4 = 9.1\,\text{N}.
\]
This force, at $24\,\si{\hertz}$, is what the plinth still feels through
a rigid baseplate, and it is what the second half of the job removes.

\subsection{Sizing the isolation mount}

The pump and baseplate, total isolated mass $M = 320\,\si{\kilogram}$
(rotor plus casing plus baseplate), are to be remounted on isolators
that pass no more than ten percent of the residual $24\,\si{\hertz}$
force to the plinth. The design follows the transmissibility curve
directly.

\textbf{Target frequency ratio.} For a transmissibility of $0.10$ with
light damping, $\mathrm{TR}\approx 1/(r^2 - 1) = 0.10$ gives $r^2 = 11$,
$r = 3.32$. We carry a margin and aim for $r = 4$, the cleaner working
choice, which sets the mount natural frequency at
\[
f_n = \frac{f_{\text{run}}}{r} = \frac{24}{4} = 6.0\,\si{\hertz}.
\]

\textbf{Stiffness.} $\omega_n = 2\pi\cdot 6 = 37.7\,\si{\radian\per
\second}$, so the total mount stiffness is
\[
k = M\omega_n^2 = 320\times (37.7)^2 = 4.55\times 10^5\,\si{\newton\per
\meter},
\]
distributed over four mounts at $1.14\times 10^5\,\si{\newton\per
\meter}$ each.

\textbf{Static deflection check.} A low mount frequency means a soft
mount and a visible static sag. The static deflection under the
machine's weight is
\[
\delta_{\text{st}} = \frac{Mg}{k} = \frac{320\times 9.81}{4.55\times
10^5} = 6.9\,\si{\milli\meter},
\]
which also recovers the mount frequency by the handy rule $f_n \approx
(1/2\pi)\sqrt{g/\delta_{\text{st}}} = (1/2\pi)\sqrt{9.81/0.0069} =
6.0\,\si{\hertz}$. A seven-millimetre sag is acceptable for a pump on
flexible piping connections; a stiffer service would force a higher
$f_n$ and a worse isolation.

\textbf{Damping and the startup peak.} Choose $\zeta = 0.07$, enough to
cap the resonance the machine crosses on every start without spoiling
the high-frequency rolloff. The verification curve is
Figure~\ref{fig:vol03ch06:casestudy-isolation}.

\input{volumes/03-force/ch06-vibrations/figures/fig-casestudy-isolation.tex}

\subsection{Verification}

The transmissibility at the running speed, $r = 4$, $\zeta = 0.07$, is
\[
\mathrm{TR} = \sqrt{\frac{1 + (2\cdot 0.07\cdot 4)^2}{(1 - 16)^2 +
(2\cdot 0.07\cdot 4)^2}} = \sqrt{\frac{1 + 0.314}{225 + 0.314}} =
\sqrt{\frac{1.314}{225.3}} = 0.0764,
\]
so the transmitted force is $0.0764\times 9.1 = 0.70\,\text{N}$, under
the ten-percent target and a factor of thirteen below the unmounted
force. The startup peak at resonance is $1/(2\zeta) = 7.1$, so during
spin-up through $6\,\si{\hertz}$ the machine briefly amplifies the
disturbance by seven; a soft-start that walks the speed through
resonance in under a second keeps the transient brief. At the residual
$2\times$ order, $48\,\si{\hertz}$ ($r = 8$), the transmissibility is
$0.030$, so any misalignment that creeps back is also held down.

The full arc closes here. A measurement diagnosed a misalignment that a
single-number overall-level alarm would have missed; the mechanical
fix removed the dominant fault; the isolator handled the residual that
cannot be removed; and the same transmissibility formula that sized the
mount verified it. The discipline is the chain, not any one formula: a
spectrum read against the fault table, a fix matched to the mechanism,
and an isolator sized and checked against a transmitted-force budget.
The reference numbers reproduce with \path{code/transmissibility.py}
for the mount and \path{code/fft_modal.py} for the spectrum.

\section{Case study: tuning a quarter-car suspension}\pathtag{standard}

The machinery case study sized a mount to keep one disturbance
frequency out of a foundation. A vehicle suspension poses the harder
version of the same problem: a base excitation spread across a band, two
masses rather than one, and a comfort target that fights a road-holding
target through the single knob of suspension damping. We carry one
corner of a passenger car from a ride-frequency target through the
spring and damper sizing to a check against the road-input spectrum, and
read off the compromise that every suspension engineer settles.

\subsection{The quarter-car model and the design targets}

One corner of the car reduces to the two-mass quarter-car of
Figure~\ref{fig:vol03ch06:quarter-car}. The sprung mass $m_s$ is a
quarter of the body; the unsprung mass $m_u$ is the wheel, tyre, brake,
and the outboard half of the suspension links. The suspension spring
$k_s$ and damper $c_s$ connect them, and the tyre stiffness $k_t$
couples the wheel to the road profile $u(t)$.

\input{volumes/03-force/ch06-vibrations/figures/fig-quarter-car.tex}

Take a mid-size car
corner: $m_s = 320\,\si{\kilogram}$, $m_u = 40\,\si{\kilogram}$, tyre
stiffness $k_t = 2.0\times 10^5\,\si{\newton\per\meter}$.

Two targets drive the design. Ride comfort wants the body-bounce
natural frequency low, in the band $1.0$ to $1.5\,\si{\hertz}$, because
the human body tolerates vertical acceleration least around $4$ to
$8\,\si{\hertz}$ and a soft suspension pushes the body resonance well
below that band. Road holding wants the tyre kept in firm contact with
the road, which limits how soft and how lightly damped the suspension
can be before the wheel-hop mode lets the tyre lift. We target a
body-bounce frequency $f_s = 1.2\,\si{\hertz}$, a deliberate choice at
the comfortable end.

\subsection{Sizing the spring}

The body-bounce mode, with the wheel held still, is a SDOF oscillator of
mass $m_s$ on the suspension spring $k_s$ in series with the tyre. The
tyre is far stiffer than the suspension ($k_t \gg k_s$), so the series
combination is close to $k_s$ alone and the body frequency is
$f_s \approx (1/2\pi)\sqrt{k_s/m_s}$. Solving for the spring,
\[
k_s = m_s(2\pi f_s)^2 = 320\times (2\pi\cdot 1.2)^2 = 320\times 56.8 =
1.82\times 10^4\,\si{\newton\per\meter}.
\]
The static deflection under the sprung weight is
$\delta_{\text{st}} = m_s g/k_s = 320\times 9.81/1.82\times 10^4 =
0.173\,\si{\meter}$, a $173\,\si{\milli\meter}$ sag that the spring's
free length and the suspension travel must accommodate, the reason a
soft ride needs long-travel suspension.

\subsection{The two modes and the wheel-hop frequency}

The quarter-car carries two natural frequencies. The low one is the
body-bounce mode at the $1.2\,\si{\hertz}$ we tuned. The high one is the
wheel-hop mode, in which the unsprung mass bounces on the tyre stiffness
with the body nearly stationary,
\[
f_u \approx \frac{1}{2\pi}\sqrt{\frac{k_s + k_t}{m_u}} = \frac{1}{2\pi}
\sqrt{\frac{1.82\times 10^4 + 2.0\times 10^5}{40}} = \frac{1}{2\pi}
\sqrt{5.46\times 10^3} = 11.8\,\si{\hertz}.
\]
The two modes sit a decade apart, $1.2$ and $11.8\,\si{\hertz}$, which
is what lets a single damper address both: the suspension damper is
sized for the body mode, and the same damper provides enough wheel-hop
control because the tyre carries little inherent damping of its own.

\subsection{Choosing the damping}

The damping ratio of the body mode sets the ride-handling balance, and
the transmissibility curve of Figure~\ref{fig:vol03ch06:quarter-car-transmissibility}
shows why no single value wins. A soft damping ratio near
$\zeta = 0.15$ gives the gentlest ride above the body resonance but a
tall peak at $1.2\,\si{\hertz}$ that the body crosses over every long
swell in the road, producing float and motion sickness. A firm
$\zeta = 0.7$ caps that peak and keeps the tyre planted for cornering
and braking, at the cost of transmitting road texture harshly into the
cabin. Production road cars settle near $\zeta = 0.25$ to $0.35$; we
pick $\zeta = 0.30$.

\input{volumes/03-force/ch06-vibrations/figures/fig-quarter-car-transmissibility.tex}

The suspension damping coefficient follows from the body mode:
\[
c_s = 2\zeta\sqrt{k_s m_s} = 2\times 0.30\times\sqrt{1.82\times 10^4
\times 320} = 0.6\times 2.41\times 10^3 = 1.45\times 10^3\,\si{\newton
\second\per\meter}.
\]
This is the damper rate the shock absorber must deliver around the body
frequency. Real dampers are deliberately asymmetric, firmer in rebound
than in compression, so the single linear $c_s$ is the equivalent rate
that reproduces the body-mode decay; the asymmetry is a refinement on
top of it.

\subsection{Checking the ride against the road spectrum}

Road roughness is a random base input whose displacement power spectral
density falls steeply with spatial frequency, so the energy the car
meets at a given temporal frequency depends on speed. At a cruising
speed the dominant body-mode excitation lands right at the $1.2\,\si{
\hertz}$ resonance, which is exactly why the resonant peak there is the
quantity the damper must control. At $\zeta = 0.30$, the body
transmissibility at resonance is
\[
\mathrm{TR}_{\text{peak}} = \frac{1}{2\zeta}\sqrt{1 + (2\zeta)^2} =
\frac{1}{0.60}\sqrt{1 + 0.36} = 1.94,
\]
so a road undulation that drives the wheel at $1.2\,\si{\hertz}$ is
amplified at the body by about a factor of two, against the factor of
$1/(2\zeta) = 6.7$ that the soft $\zeta = 0.15$ setting would have let
through. Above $r = \sqrt2$, that is above roughly $1.7\,\si{\hertz}$,
all three damping curves drop below unity and the suspension isolates;
the firmer damper isolates less well there, the harshness penalty for
the better resonance control. The wheel-hop mode at $11.8\,\si{\hertz}$
gets its own small transmissibility peak, but at $\zeta = 0.30$ referred
to the body mode the wheel-hop peak stays modest and the tyre keeps
contact.

\subsection{The compromise, stated}

The corner is tuned: $k_s = 1.82\times 10^4\,\si{\newton\per\meter}$
gives a $1.2\,\si{\hertz}$ ride frequency, $c_s = 1.45\times 10^3\,\si{
\newton\second\per\meter}$ gives $\zeta = 0.30$, and the wheel-hop mode
sits cleanly at $11.8\,\si{\hertz}$. The body resonance is held to a
transmissibility near $2$ rather than the $6.7$ a comfort-first damper
would allow, and the high-frequency ride is firmer than that
comfort-first damper by the same trade. A passive suspension cannot beat
the curve crossing at $r = \sqrt2$: every gain in resonance control is
paid in high-frequency harshness, and the reverse. The escape is to make
$c_s$ depend on the road, which is what semi-active dampers
(magnetorheological or solenoid-valved) and fully active suspensions do,
softening for the smooth high-frequency texture and firming for the
body-mode swell. The same transmissibility algebra that sized this
corner is the cost function those controllers minimise, and
\path{code/transmissibility.py} tabulates the body curve for any
$\zeta$ the reader wants to compare.

\section{Failure: Tacoma Narrows, Millennium Bridge, machinery
resonance}\pathtag{core}

Vibration failures share a structure: a system has a lightly-damped
natural mode at a frequency near the dominant excitation, and the
amplification at resonance produces motion the structure cannot
sustain. The chapter's failure section names the canonical cases.

\subsection{Tacoma Narrows: aeroelastic flutter, revisited}

Chapter 5 of this volume introduced the original Tacoma Narrows
Bridge collapse of 7 November 1940 as the canonical case of a
missing degree of freedom in an analytical-mechanics analysis. The
vibration perspective on the same event is that the bridge had a
lightly-damped torsional natural mode whose interaction with a
steady wind produced negative aerodynamic damping above a critical
wind speed of approximately $19\,\si{\meter\per\second}$; the
effective total damping became negative; the torsional vibration
grew exponentially until the deck's structural capacity was exceeded
\cite{acc:tacoma-narrows-1941}.

The aerodynamic-damping mechanism is the key. The deck section in
steady wind experiences a pitching moment that depends on the angle
of attack. When the deck oscillates torsionally, the angle of attack
changes; the resulting pitching moment has both a component in phase
with the angular displacement (an aerodynamic stiffness) and a
component in phase with the angular velocity (an aerodynamic
damping). At low wind speeds, the aerodynamic damping is positive
and adds to the structural damping. Above a critical wind speed, the
aerodynamic damping turns negative; at the critical speed, the
total damping (structural plus aerodynamic) reaches zero; above the
critical speed, the total damping is negative, and the natural mode
becomes self-excited.

In the engineering-vibration framework of this chapter, Tacoma
Narrows is a case of a structurally lightly-damped natural mode
($\zeta \approx 0.005$, typical of welded-steel bridge structures)
that loses its damping to an aerodynamic feedback term. The
remediation in subsequent bridge designs is either higher structural
damping (truss-stiffened decks add internal friction at the truss
joints) or aerodynamic-section shapes whose damping does not turn
negative (streamlined deck sections used on modern long-span
bridges).

\subsection{Millennium Bridge: synchronous lateral excitation}

The London Millennium Footbridge across the Thames opened on 10
June 2000 and was closed two days later, on 12 June 2000, when
crowds of pedestrians produced visible lateral oscillations of the
deck of approximately $70\,\si{\milli\meter}$ amplitude at
frequencies near $1\,\si{\hertz}$. The mechanism is
\engterm{synchronous lateral excitation}: pedestrians instinctively
adjust their gait to maintain balance when the deck moves laterally,
locking their per-step lateral footfall forces in phase with the
deck's lateral velocity and feeding energy into the bridge's
lateral mode. Above a critical pedestrian density of approximately
$1.3\,\text{persons}/\si{\meter\squared}$, the coherent forcing
exceeds the deck's structural damping and the lateral mode becomes
self-excited \cite{acc:dallard-millennium-2001}.

The mechanism is novel in two ways. First, the forcing is itself a
function of the motion, like the aerodynamic feedback in Tacoma
Narrows, but generated by human balance reflexes rather than
fluid-structure interaction. Second, the random-phase
incoherent-forcing model used in classical pedestrian-load analysis
(which underestimates the response by $\sim\sqrt{N}$ rather than
$\sim N$ in the pedestrian count) does not apply once the
synchronisation sets in. The Arup engineering team led by Pat Dallard
identified the mechanism in the year following the closure and
remediated the bridge by installing $26$ fluid-viscous dampers and
$52$ tuned mass dampers, adding approximately $20\,\%$ critical
damping to the lateral modes and moving the critical pedestrian
density above any plausible loading. The bridge reopened on 22
February 2002.

In the engineering-vibration framework, the Millennium Bridge is a
case of a structurally lightly-damped natural mode whose damping is
overcome by a load-coupled feedback term. The same diagnostic
principle as Tacoma Narrows applies: an apparently inadequately
analysed structure exhibits a vibration failure mode in service that
the design analysis did not include. The structural-vibration
analysis was correct for the mode it represented; the analysis did
not represent the mode that was excited.

\subsection{Machinery resonance: less famous but more common}

Industrial machinery (pumps, fans, compressors, turbines, motors,
generators) almost always has some component whose natural
frequency falls in the operating-speed band of the machine. When the
operating speed coincides with the component's natural frequency,
the component's response is amplified by $1/(2\zeta)$ and may
exceed its design capacity. The result is rapid wear, fatigue
failure of bolts or shafts, audible drumming, or in extreme cases
sudden fracture.

The standard machinery-resonance failure modes:

\begin{itemize}[leftmargin=*]
  \item \engterm{Critical speed of rotating machinery}. A spinning
    shaft has bending natural frequencies (whirl modes); when the
    rotation speed coincides with a whirl natural frequency, the
    shaft's centroid orbits the geometric axis at large amplitude.
    Designers either keep the operating speed below the first
    critical (subcritical operation, conservative but heavy) or
    accelerate quickly through the criticals to a supercritical
    operating point (lighter, but transient critical-speed crossings
    must be managed).
  \item \engterm{Piping resonance}. Process plant piping vibrates
    when the fluid pulsation frequency (from positive-displacement
    pumps, compressors, or flow-induced turbulence) coincides with
    the piping's mechanical natural frequency. The vibration
    fatigues the piping at supports and at branch connections, often
    leaking after months or years rather than failing catastrophically.
  \item \engterm{Foundation resonance}. A machine's foundation,
    treated as a rigid structure in much design analysis, has its
    own modes. When a machine's vibration excites a foundation mode,
    the machine-foundation system's response can exceed the
    machine-on-stiff-foundation prediction by factors of $10$ or
    more. The mitigation is foundation stiffening, foundation
    isolation, or operating-speed change.
  \item \engterm{Blade-pass resonance in turbomachinery}. A rotor
    blade passes a stator vane at a frequency equal to (rotor speed)
    $\times$ (number of stator vanes). If this frequency coincides
    with a blade's natural frequency, the blade vibrates at large
    amplitude and fatigues at its hub fillet. Design discipline is
    to detune the blade so the resonance falls outside the operating
    range.
\end{itemize}

The remediation in every case is the same: change the natural
frequency (stiffness or mass modification), change the driving
frequency (operating-speed change), or add damping. The choice
depends on the cost and the design constraints.

\begin{principle}[Vibration failure is a frequency-coincidence
failure]
A vibration-related structural or machinery failure occurs when (1)
a lightly-damped natural mode of the system coincides in frequency
with (2) a driving excitation, and (3) the resulting resonance
amplification exceeds the structure's capacity. Mitigating any one
of the three (moving the natural frequency, removing or detuning
the excitation, adding damping, or strengthening the structure)
prevents the failure. The design discipline is to identify the
natural frequencies, the expected excitation spectrum, the damping,
and to verify that no coincidence in the operating envelope
produces a response above capacity.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Characterise the vibration response of a household
object]
The reader instruments a household structural object (a metre stick
clamped as a cantilever, a wooden beam supported as a simply-
supported member, a metal ruler, a guitar string, or similar),
measures its frequency response under impact excitation, extracts
the lowest natural frequency and the damping ratio, and compares
to a beam- or string-vibration theoretical prediction. The
deliverable is a twenty- to thirty-page engineering report.

\textbf{Track}. Physical measurement plus analysis. \textbf{Hazard
class}: none. Impact excitation by a light mallet or by snapping the
finger against the object.

\textbf{Apparatus}. A measurement device: a smartphone with a free
accelerometer-logging app (most smartphones have built-in
accelerometers logged at $50$ to $200\,\si{\hertz}$), or a contact
microphone attached to the object, or a strain-gauge bridge as in
the Chapter 3 project. The reader's chosen vibrating object. A
clamp or support fixture. A timing reference (the smartphone's
clock).

\textbf{Procedure}. The reader sets up the vibrating object as a
known geometric configuration (cantilever, simply supported, free-
free), excites it by a light impact at one end, and records the
time history of the response. The reader records at least five
different excitation locations and at least three different impact
strengths.

\textbf{Analysis}. From each recorded time history, the reader:

\begin{itemize}[leftmargin=*]
  \item Estimates the fundamental natural period from the time
    between successive zero-crossings of the response.
  \item Computes the natural frequency in Hz.
  \item Estimates the damping ratio from the exponential decay of
    successive amplitude peaks (the logarithmic-decrement method).
  \item Computes the FFT of the recorded signal and identifies the
    first three peaks; reports their frequencies and the ratios
    among them.
\end{itemize}

\textbf{Theoretical prediction}. The reader computes the
theoretical fundamental natural frequency from the object's geometry
and material properties using the cantilever-beam formula, the
simply-supported-beam formula, or the string formula as
appropriate.

\textbf{Comparison}. The reader compares the measured fundamental
frequency to the predicted fundamental frequency and discusses
sources of discrepancy.

\textbf{Reflection ($1{,}500$ to $3{,}000$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
  \item Which sources of damping (material hysteresis, joint
    friction, air drag, acoustic radiation) dominate at the
    measured damping ratio.
  \item Whether the higher-frequency peaks in the FFT match the
    theoretical higher modes (in their ratios) or are something
    else (room acoustics, sensor artifacts, electrical noise).
  \item One feature of the measurement (resolution, accelerometer
    noise floor, finite record length) that limits the experiment's
    accuracy.
  \item One quantitative prediction (next mode's natural frequency,
    response to a different boundary condition) the reader could
    make from the measured data and verify with a second
    experiment.
\end{itemize}

\textbf{Deliverable}. The recorded time histories, the FFT plots,
the natural-frequency and damping estimates, the theoretical
predictions, the comparison, and the reflection.

\textbf{Time}. Two to four hours of measurement, four to eight hours
of analysis (FFT, log-decrement extraction), four to eight hours of
write-up.
\end{project}

\section{Exercises}

The exercises mix calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A mass of $2\,\si{\kilogram}$ on a spring of stiffness $200\,\si{
\newton\per\meter}$ has no damping. Find the natural frequency in Hz
and the period.

\paragraph{Solution sketch.} $\omega_n = \sqrt{k/m} = \sqrt{200/2} =
10\,\si{\radian\per\second}$. Then $f_n = \omega_n/(2\pi) = 1.59\,\si{
\hertz}$ and $T_n = 1/f_n = 0.628\,\si{\second}$.
\end{exercise}

\begin{exercise}
A pendulum has period $T = 1.5\,\si{\second}$ for small swings. Find
its length, assuming $g = 9.81\,\si{\meter\per\second\squared}$.

\paragraph{Solution sketch.} From $T = 2\pi\sqrt{L/g}$, solve for
$L = g(T/2\pi)^2 = 9.81 \times (1.5/6.283)^2 = 9.81 \times 0.0570 =
0.559\,\si{\meter}$. A pendulum a little over half a metre long
swings once every $1.5\,\si{\second}$.
\end{exercise}

\begin{exercise}
A SDOF system has $m = 5\,\si{\kilogram}$, $k = 500\,\si{\newton\per
\meter}$, and damping ratio $\zeta = 0.05$. Find the damped natural
frequency. State the damping coefficient $c$.

\paragraph{Solution sketch.} $\omega_n = \sqrt{500/5} = 10\,\si{
\radian\per\second}$. The damped frequency is $\omega_d = \omega_n
\sqrt{1 - \zeta^2} = 10\sqrt{1 - 0.0025} = 9.9875\,\si{\radian\per
\second}$, so $f_d = 1.589\,\si{\hertz}$. The damping coefficient is
$c = 2\zeta m \omega_n = 2 \times 0.05 \times 5 \times 10 = 5.0\,\si{
\newton\second\per\meter}$. The damping shifts the frequency by only
$0.125\,\%$, the usual situation for $\zeta = 0.05$.
\end{exercise}

\begin{exercise}
A SDOF system with natural frequency $\omega_n = 10\,\si{\radian\per
\second}$ and damping ratio $\zeta = 0.02$ is driven at $\omega =
\omega_n$ by a force of amplitude $F_0 = 50\,\text{N}$. Find the
steady-state response amplitude.

\paragraph{Solution sketch.} The stiffness is $k = m\omega_n^2$, but
$m$ is not given separately, so leave the answer in terms of the
amplification of the static deflection. At resonance $X =
(F_0/k)/(2\zeta)$. With $k = m\omega_n^2$, $X = F_0/(2\zeta m
\omega_n^2)$. If the system is taken to have $m = 1\,\si{\kilogram}$
(so $k = 100\,\si{\newton\per\meter}$), then $F_0/k = 0.50\,\si{
\meter}$ and $X = 0.50/(2 \times 0.02) = 12.5\,\si{\meter}$, an
amplification of $1/(2\zeta) = 25$. The point is the factor $25$; the
absolute amplitude scales with the assumed mass.
\end{exercise}

\begin{exercise}
A vibration isolator has natural frequency $f_n = 5\,\si{\hertz}$.
Find the transmissibility at a driving frequency of $20\,\si{
\hertz}$, with $\zeta = 0.1$.

\paragraph{Solution sketch.} The frequency ratio is $r = 20/5 = 4$.
Then
\[
\mathrm{TR} = \sqrt{\frac{1 + (2\cdot 0.1\cdot 4)^2}{(1 - 16)^2 +
(2\cdot 0.1\cdot 4)^2}} = \sqrt{\frac{1 + 0.64}{225 + 0.64}} =
\sqrt{\frac{1.64}{225.64}} = 0.0853.
\]
The mount transmits about $8.5\,\%$ of the support motion. The
light-damping approximation $1/(r^2 - 1) = 1/15 = 0.067$ is lower
because it omits the damping term that limits high-frequency
isolation.
\end{exercise}

\begin{exercise}
A simply-supported beam of length $4\,\si{\meter}$, density $\rho =
7{,}800\,\si{\kilogram\per\meter\cubed}$, cross-section $50\,\si{
\milli\meter} \times 25\,\si{\milli\meter}$ (depth direction
$25\,\si{\milli\meter}$), and $E = 200\,\text{GPa}$. Find the first
three natural frequencies.

\paragraph{Solution sketch.} Section properties: $A = 0.050 \times
0.025 = 1.25 \times 10^{-3}\,\si{\meter\squared}$, $I = bh^3/12 =
0.050 \times (0.025)^3/12 = 6.51 \times 10^{-8}\,\si{\meter\tothe{4}}$.
Then $EI = 200 \times 10^9 \times 6.51 \times 10^{-8} = 1.302 \times
10^4\,\si{\newton\meter\squared}$ and $\rho A = 7800 \times 1.25
\times 10^{-3} = 9.75\,\si{\kilogram\per\meter}$. For a simply
supported beam $\omega_n = (n\pi)^2\sqrt{EI/(\rho A L^4)}$ with $L =
4\,\si{\meter}$, so the bracket is $\sqrt{1.302\times 10^4/(9.75
\times 256)} = \sqrt{5.215} = 2.284$. Then $\omega_1 = \pi^2 \times
2.284 = 22.5\,\si{\radian\per\second}$ ($f_1 = 3.59\,\si{\hertz}$),
$\omega_2 = 4\omega_1 = 90.2\,\si{\radian\per\second}$ ($f_2 = 14.3\,
\si{\hertz}$), $\omega_3 = 9\omega_1 = 203\,\si{\radian\per\second}$
($f_3 = 32.3\,\si{\hertz}$). The frequencies follow the $1:4:9$ ratio
of a simply supported beam.
\end{exercise}

\begin{exercise}
A guitar string of length $0.65\,\si{\meter}$ and mass per unit length
$5\,\si{\gram\per\meter}$ has tension $T = 100\,\text{N}$. Find its
fundamental frequency and the next two harmonics.

\paragraph{Solution sketch.} The wave speed is $c = \sqrt{T/\rho} =
\sqrt{100/0.005} = 141.4\,\si{\meter\per\second}$. The fundamental is
$f_1 = c/(2L) = 141.4/(2 \times 0.65) = 108.8\,\si{\hertz}$, near
concert A. The harmonics are integer multiples: $f_2 = 217.6\,\si{
\hertz}$ and $f_3 = 326.4\,\si{\hertz}$. The harmonic series is exact
for the ideal string; a real string's finite bending stiffness sharps
the upper harmonics slightly.
\end{exercise}

\begin{exercise}\label{ex:vol03ch06:2dof}
A two-degree-of-freedom system has $M = \text{diag}(2, 1)\,\si{
\kilogram}$ and $K$ matrix $\begin{pmatrix} 3 & -1 \\ -1 & 1
\end{pmatrix}\,\si{\newton\per\meter}$. Find the two natural
frequencies and the corresponding mode shapes.

\paragraph{Solution sketch.} Worked in full in the Worked-examples
section above. The characteristic equation $\det(K - \omega^2 M) =
2\omega^4 - 5\omega^2 + 2 = 0$ gives $\omega^2 = 0.5$ and $2$, so
$\omega_1 = 0.707\,\si{\radian\per\second}$ and $\omega_2 = 1.414\,\si{
\radian\per\second}$. The mode shapes are $\vec{\phi}_1 = [1, 2]^T$
(in phase) and $\vec{\phi}_2 = [1, -1]^T$ (out of phase). The program
\path{code/modal_2dof.py} reproduces these and verifies
mass-orthogonality.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the equation of motion of a SDOF system from Newton's second
law for a mass on a horizontal spring. State the assumption about
the spring's linearity and the meaning of the displacement variable.

\paragraph{Solution sketch.} Let $x$ be the displacement of the mass
from the spring's natural (unstretched) length, positive in the
stretching direction. A linear spring exerts a restoring force
$-kx$ proportional to and opposing the displacement; this is the
linearity assumption, valid for small displacements where the
force-extension curve is straight. Newton's second law along the line
of motion reads $m\ddot{x} = -kx$, that is $m\ddot{x} + kx = 0$. The
assumption fails once the displacement is large enough that the spring
softens or stiffens, at which point the equation gains nonlinear
terms.
\end{exercise}

\begin{exercise}
Derive the natural frequency $\omega_n = \sqrt{g/L}$ for a simple
pendulum at small amplitude, starting from the work-energy theorem
or the Lagrangian.

\paragraph{Solution sketch.} With swing angle $\theta$ from the
vertical, the bob's kinetic energy is $T = \tfrac{1}{2}mL^2\dot
\theta^2$ and its potential energy is $U = mgL(1 - \cos\theta)$. The
Lagrangian $\mathcal{L} = T - U$ gives the Euler-Lagrange equation
$mL^2\ddot\theta + mgL\sin\theta = 0$, that is $\ddot\theta +
(g/L)\sin\theta = 0$. For small amplitude $\sin\theta \approx \theta$,
so $\ddot\theta + (g/L)\theta = 0$, a simple harmonic oscillator with
$\omega_n = \sqrt{g/L}$. The mass cancels: the pendulum's frequency
depends only on length and gravity.
\end{exercise}

\begin{exercise}
Derive the logarithmic decrement $\delta = 2\pi\zeta/\sqrt{1-\zeta^2}$
from the underdamped solution $x(t) = X_0 e^{-\zeta\omega_n t}\cos(
\omega_d t - \phi)$.

\paragraph{Solution sketch.} Two successive peaks one damped period
$T_d = 2\pi/\omega_d$ apart have amplitudes set by the envelope
$X_0 e^{-\zeta\omega_n t}$ at $t$ and $t + T_d$. Their ratio is
\[
\frac{x(t)}{x(t + T_d)} = \frac{e^{-\zeta\omega_n t}}{e^{-\zeta
\omega_n (t + T_d)}} = e^{\zeta\omega_n T_d}.
\]
Taking the logarithm, $\delta = \zeta\omega_n T_d = \zeta\omega_n
\cdot 2\pi/\omega_d$. Since $\omega_d = \omega_n\sqrt{1 - \zeta^2}$,
the factor $\omega_n/\omega_d = 1/\sqrt{1 - \zeta^2}$, so $\delta =
2\pi\zeta/\sqrt{1 - \zeta^2}$. For light damping the radical is near
unity and $\delta \approx 2\pi\zeta$.
\end{exercise}

\begin{exercise}
Derive the resonance amplification factor $1/(2\zeta)$ for a
forced SDOF system at $\omega = \omega_n$, starting from the
frequency response function $X(\omega)/(F_0/k)$.

\paragraph{Solution sketch.} The amplification is
\[
\frac{X}{F_0/k} = \frac{1}{\sqrt{(1 - r^2)^2 + (2\zeta r)^2}}, \quad
r = \frac{\omega}{\omega_n}.
\]
At $\omega = \omega_n$, $r = 1$, so $(1 - r^2)^2 = 0$ and the radical
reduces to $\sqrt{(2\zeta)^2} = 2\zeta$. Therefore the amplification
is $1/(2\zeta)$. (The true maximum sits at $r = \sqrt{1 - 2\zeta^2}$
with value $1/(2\zeta\sqrt{1-\zeta^2})$, but for light damping the
$r = 1$ result is the standard working figure.)
\end{exercise}

\begin{exercise}
Derive the natural frequencies of a string fixed at both ends from
the wave equation $\partial^2 y/\partial t^2 = c^2 \partial^2 y/
\partial x^2$ with boundary conditions $y(0, t) = y(L, t) = 0$.

\paragraph{Solution sketch.} Separate variables, $y(x, t) =
\Phi(x)q(t)$. Substitution gives $\Phi q'' = c^2\Phi'' q$, that is
$q''/q = c^2\Phi''/\Phi = -\omega^2$ for a separation constant. The
spatial equation $\Phi'' + (\omega/c)^2\Phi = 0$ has solution $\Phi =
A\sin(\omega x/c) + B\cos(\omega x/c)$. The condition $\Phi(0) = 0$
forces $B = 0$; the condition $\Phi(L) = 0$ forces $\sin(\omega L/c)
= 0$, so $\omega L/c = n\pi$ for integer $n \geq 1$. Therefore
$\omega_n = n\pi c/L = (n\pi/L)\sqrt{T/\rho}$, the harmonic series.
The mode shapes are $\Phi_n(x) = \sin(n\pi x/L)$, shown in
Figure~\ref{fig:vol03ch06:mode-shapes}.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the natural frequency of a chair when shaken side-to-side
by a sitting person. State the dominant mass (the chair, the
person, both) and the stiffness mechanism.

\paragraph{Reference estimate.} The moving mass is the person plus the
chair seat, on the order of $80\,\si{\kilogram}$. The stiffness is the
lateral bending stiffness of the chair legs acting as short
cantilevers; a wooden or metal leg set might give an effective lateral
stiffness of order $10^4$ to $10^5\,\si{\newton\per\meter}$. Then
$f_n = (1/2\pi)\sqrt{k/m}$ ranges from $(1/2\pi)\sqrt{10^4/80} =
1.8\,\si{\hertz}$ to $(1/2\pi)\sqrt{10^5/80} = 5.6\,\si{\hertz}$. A
chair sways at a few hertz, which is why a person rocking a chair
feels a slow wobble rather than a buzz. The dominant mass is the
person; the stiffness is the legs in lateral bending.
\end{exercise}

\begin{exercise}
Estimate the fundamental natural frequency of a residential wood
floor joist of $4\,\si{\meter}$ span carrying a uniform live load
of $2\,\text{kN}/\text{m}$. State assumptions about the joist's
cross-section and material; compare to the typical resonant
frequency band of walking.

\paragraph{Reference estimate.} Assume a $38 \times 235\,\si{\milli
\meter}$ joist ($2\times 10$ lumber), $E \approx 10\,\text{GPa}$. Then
$I = bh^3/12 = 0.038 \times (0.235)^3/12 = 4.11 \times 10^{-5}\,\si{
\meter\tothe{4}}$ and $EI = 4.11 \times 10^5\,\si{\newton\meter
\squared}$. The live load $2\,\text{kN}/\text{m}$ contributes a mass
per length of $2000/9.81 = 204\,\si{\kilogram\per\meter}$; add a few
kilograms for the joist itself, $\rho A \approx 208\,\si{\kilogram\per
\meter}$. For a simply supported beam $f_1 = (\pi/2)\sqrt{EI/(\rho A
L^4)} = (\pi/2)\sqrt{4.11\times 10^5/(208 \times 256)} = 4.4\,\si{
\hertz}$. Walking forces concentrate near $1.5$ to $2.5\,\si{\hertz}$
with harmonics at $3$ to $5\,\si{\hertz}$, so a $4.4\,\si{\hertz}$
joist sits uncomfortably close to the second walking harmonic, the
reason residential floor-vibration codes target a fundamental above
roughly $8\,\si{\hertz}$ for lively floors.
\end{exercise}

\begin{exercise}
Estimate the damping ratio of a swing set in a park, measured by
the rate at which a swing's amplitude decays after the rider stops
pumping. State the dominant damping source.

\paragraph{Reference estimate.} A playground swing of length $L
\approx 2.5\,\si{\meter}$ has period $T = 2\pi\sqrt{L/g} = 3.2\,\si{
\second}$. A rider who stops pumping typically takes on the order of
$20$ to $30$ swings to decay to a small fraction of the starting
amplitude. If the amplitude halves every ten swings, $\delta =
(1/10)\ln 2 = 0.069$ and $\zeta \approx \delta/(2\pi) = 0.011$. The
damping ratio is around $1\,\%$, very light. The dominant source is
aerodynamic drag on the rider's body (which scales with velocity
squared, so the decay is faster at high amplitude), with a smaller
contribution from friction at the chain pivots. A bare swing with no
rider decays more slowly still.
\end{exercise}

\begin{exercise}
Estimate the natural frequency of a tall office building's first
sway mode. State assumptions about the building's height, mass per
unit height, and effective stiffness.

\paragraph{Reference estimate.} The fastest route is the
structural-engineering rule of thumb $T_1 \approx 0.1 N\,\si{\second}$
for an $N$-storey building, which encodes the empirical fact that
taller buildings are not proportionally stiffer. A twenty-storey
building has $T_1 \approx 2\,\si{\second}$, $f_1 \approx 0.5\,\si{
\hertz}$; a forty-storey building has $T_1 \approx 4\,\si{\second}$,
$f_1 \approx 0.25\,\si{\hertz}$. A cantilever-beam model with height
$H$, mass per height $\rho A$, and bending stiffness $EI$ gives $f_1 =
(1.875)^2/(2\pi)\sqrt{EI/(\rho A H^4)}$ and reproduces the same band
for plausible steel-frame stiffness. The lower the frequency, the more
the building responds to the low-frequency end of the wind-gust
spectrum, which is why supertall towers carry tuned mass dampers.
\end{exercise}

\begin{exercise}
Estimate the natural frequency of a soda bottle when partially
filled with water and tapped at the rim. State the analogy with a
Helmholtz resonator and what happens to the frequency as the water
level rises.

\paragraph{Reference estimate.} Blowing across the rim (the cleaner
excitation than tapping) drives a Helmholtz resonator: the air in the
neck is the moving mass, the air in the body is the spring. The
frequency is $f = (c/2\pi)\sqrt{A/(V L_{\text{eff}})}$, with $c =
343\,\si{\meter\per\second}$ the speed of sound, $A$ the neck area,
$V$ the air-cavity volume, and $L_{\text{eff}}$ the effective neck
length. For a $0.5\,\si{\liter}$ bottle with a neck of area $\sim 3\,
\si{\centi\meter\squared}$ and effective length $\sim 5\,\si{\centi
\meter}$, an air volume of $V = 0.3\,\si{\liter}$ gives $f =
(343/2\pi)\sqrt{3\times 10^{-4}/(3\times 10^{-4} \times 0.05)} \approx
240\,\si{\hertz}$. As water rises, $V$ falls, so $f$ rises: the pitch
climbs as the bottle fills, the everyday demonstration of the
$f \propto V^{-1/2}$ law.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Simulate the response of an undamped SDOF system to a sinusoidal
forcing at $\omega/\omega_n = 0.5$, $0.99$, $1.01$, $1.5$, $2.0$.
Plot the response over twenty periods of forcing and identify the
beat frequencies for the cases close to resonance.

\paragraph{Solution sketch.} Integrate $m\ddot x + kx = F_0\cos(\omega
t)$ from rest with \path{code/sdof_response.py} (set $c = 0$). The
$r = 0.5$ and $r = 1.5$, $2.0$ cases show a steady oscillation at the
drive frequency superposed on the free response at $\omega_n$. The
$r = 0.99$ and $r = 1.01$ cases show beats: the response is a fast
oscillation at the average frequency inside a slow envelope at the
beat frequency $|\omega - \omega_n| = 0.01\,\omega_n$, with beat
period $2\pi/(0.01\,\omega_n) = 100$ drive-period-scale units. The
beat period diverges as $r \to 1$; at exactly $r = 1$ the undamped
amplitude grows linearly without bound, as in
Figure~\ref{fig:vol03ch06:beat}.
\end{exercise}

\begin{exercise}
Simulate the response of a damped SDOF system with $\zeta = 0.01$,
$0.1$, $0.5$ to a sinusoidal forcing at $\omega = \omega_n$.
Estimate the time required for each system to reach steady state.

\paragraph{Solution sketch.} The transient decays as $e^{-\zeta
\omega_n t}$, so the settling time to within $5\,\%$ is $t_s \approx
3/(\zeta\omega_n)$, that is about $3/(2\pi\zeta)$ drive periods. For
$\zeta = 0.01$ that is roughly $48$ periods of slow buildup toward the
resonant amplitude $1/(2\zeta) = 50$ times the static deflection; for
$\zeta = 0.1$, about $5$ periods toward an amplitude of $5$; for
$\zeta = 0.5$, under two periods toward an amplitude near $1$. The
simulation with \path{code/sdof_response.py} shows the lightly damped
case taking the longest to settle and reaching by far the largest
amplitude, the double penalty of light damping at resonance.
\end{exercise}

\begin{exercise}
Implement a two-degree-of-freedom modal-analysis solver. Input: mass
matrix $M$, stiffness matrix $K$. Output: natural frequencies and
mode shapes. Verify on the system of Calculation 8.

\paragraph{Solution sketch.} Solve the generalised eigenvalue problem
$K\vec\phi = \omega^2 M\vec\phi$. The reference implementation
\path{code/modal_2dof.py} calls a symmetric solver, takes the square
roots of the eigenvalues for the frequencies, mass-normalises the
eigenvectors, and checks $\vec\phi_i^T M\vec\phi_j = \delta_{ij}$. On
the system of Calculation~\ref{ex:vol03ch06:2dof} it returns
$\omega_1 = 0.707$, $\omega_2 = 1.414\,\si{\radian\per\second}$ and
mode shapes $[1, 2]^T$, $[1, -1]^T$, matching the hand calculation.
The same code, given a tridiagonal shear-frame $M$ and $K$, returns
all $n$ storey frequencies.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a passive vibration isolator for a $100\,\si{\kilogram}$
machine that vibrates at $50\,\si{\hertz}$ with $0.1\,\si{\milli
\meter}$ amplitude. Specify the isolator natural frequency, the
damping ratio, the spring stiffness, and the resulting
transmissibility.

\paragraph{Model-answer sketch.} Worked in the Worked-examples
section. Target a transmissibility around $5\,\%$. Pick an isolator
$f_n = 10.9\,\si{\hertz}$ (frequency ratio $r = 4.58$), giving total
stiffness $k = m\omega_n^2 = 4.7\times 10^5\,\si{\newton\per\meter}$,
four mounts at $118\,\si{\kilo\newton\per\meter}$ each. Choose a light
damping ratio $\zeta \approx 0.05$ so the curve rolls off cleanly above
resonance while still capping the startup peak at $1/(2\zeta) = 10$.
The resulting transmissibility at $50\,\si{\hertz}$ is $0.055$, so the
transmitted amplitude is about $0.0055\,\si{\milli\meter}$. The design
trade is the startup transient: the machine passes through
$10.9\,\si{\hertz}$ on spin-up, where it briefly amplifies.
\end{exercise}

\begin{exercise}
Design a tuned mass damper for a tall building whose lowest sway
mode is at $0.5\,\si{\hertz}$. Specify the damper's mass (as a
fraction of the building's first-mode modal mass), its natural
frequency, and its damping ratio. State the expected reduction in
the building's response amplitude.

\paragraph{Model-answer sketch.} A tuned mass damper attaches an
auxiliary mass $m_d$ on its own spring and damper to the structure.
The standard Den Hartog tuning for a mass ratio $\mu = m_d/M$ sets the
damper frequency to $f_d = f_n/(1 + \mu)$ and the damper damping ratio
to $\zeta_d = \sqrt{3\mu/[8(1 + \mu)^3]}$. Choose $\mu = 0.02$ (a
common value; the Taipei~101 damper is near this). Then $f_d =
0.5/1.02 = 0.490\,\si{\hertz}$ and $\zeta_d = \sqrt{0.06/8.49} =
0.084$. A $2\,\%$ mass ratio with optimal tuning cuts the peak
response by roughly a factor of $\sqrt{1 + 1/\mu} \approx 7$ relative
to the undamped structure, the order of reduction these dampers
deliver in practice. The damper mass for a building with $3\times
10^7\,\si{\kilogram}$ modal mass is $6\times 10^5\,\si{\kilogram}$,
several hundred tonnes.
\end{exercise}

\begin{exercise}
Design a measurement campaign to identify the first three natural
frequencies of an industrial centrifugal pump. Specify the
accelerometer placement, the excitation method, and the signal-
processing chain.

\paragraph{Model-answer sketch.} Mount triaxial accelerometers at the
bearing housings (drive and non-drive end), the casing, and the
baseplate, so that no node of the first three modes is unmeasured.
Excite with an instrumented impact hammer for a broadband impulse
(machine stopped) and, separately, run an operating-deflection-shape
measurement during a controlled run-up to separate structural
resonances from running-speed orders. Process each channel with
\path{code/fft_modal.py}: window, transform, identify the frequency
response function peaks, and fix each natural frequency by the
$90^{\circ}$ phase crossing rather than the amplitude peak. Damping
follows from the half-power bandwidth. A record length of at least
$10\,\si{\second}$ gives $0.1\,\si{\hertz}$ resolution, fine enough to
separate closely spaced modes.
\end{exercise}

\begin{exercise}
Design a vibration-monitoring threshold table for a critical
industrial fan: green/yellow/red alert levels for the $1\times$,
$2\times$, and broad-band components of the bearing vibration
signature. State the rationale for each threshold.

\paragraph{Model-answer sketch.} Express thresholds in velocity RMS
(the standard vibration-severity measure, since it weights mid-band
energy where bearing damage shows). A defensible table for a
medium machine, anchored to the ISO~10816 severity classes: green
below $2.8\,\si{\milli\meter\per\second}$ RMS, yellow $2.8$ to $7.1\,
\si{\milli\meter\per\second}$, red above $7.1\,\si{\milli\meter\per
\second}$. Apply these per-component: a rising $1\times$ flags
imbalance, a rising $2\times$ flags misalignment, and rising
broad-band energy flags bearing wear. The rationale is that absolute
level matters less than trend: a doubling of any component over a few
months is the alarm even when the absolute value is still green. A
representative baseline spectrum is in
\path{data/machine_spectrum.csv}.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
An industrial fan exhibits a vibration spectrum with a peak at
$1\times$ rotation frequency that has grown over six months. Identify
the most probable cause, the corrective action, and the maintenance
discipline that would have caught the trend earlier.

\paragraph{Solution sketch.} A growing $1\times$ peak is the
signature of rotor imbalance, here progressive: dust or product
buildup on the blades, erosion of one blade, or a loosening hub. The
corrective action is to clean and inspect the rotor and to field
balance it (add or remove trim mass until the $1\times$ component
drops). The discipline that would have caught the trend earlier is
routine condition monitoring: a monthly recorded spectrum trended over
time would have flagged the rising $1\times$ months before it reached
an alarm level, the core value of condition-based maintenance.
\end{exercise}

\begin{exercise}
A residential building exhibits an audible humming during high
winds at a frequency of $\sim 4\,\si{\hertz}$. Identify the most
probable mechanism and the verification measurement.

\paragraph{Solution sketch.} A wind-speed-dependent tone near a fixed
frequency points to vortex shedding off a slender element (a railing,
a cable, a mast, a sign frame) whose shedding frequency $f =
\mathit{St}\,U/D$ (Strouhal number $\mathit{St} \approx 0.2$, $U$ wind
speed, $D$ element diameter) has locked onto the element's mechanical
natural frequency near $4\,\si{\hertz}$. The verification measurement
is to attach an accelerometer to suspected elements during a windy
period and confirm which one resonates at $4\,\si{\hertz}$, and to
check that the tone strengthens in a wind-speed band consistent with
the Strouhal relation. The fix is to detune or add damping to the
offending element, or to fit a helical strake that disrupts coherent
shedding.
\end{exercise}

\begin{exercise}
A pedestrian bridge exhibits visible vertical oscillation at
$\sim 2\,\si{\hertz}$ when groups of pedestrians cross. Identify
whether this is a synchronous-vertical-excitation phenomenon
analogous to the Millennium Bridge's lateral mode, or a different
mechanism, and state the diagnostic measurement.

\paragraph{Solution sketch.} The walking pace is about $2\,\si{
\hertz}$, so a vertical mode at $2\,\si{\hertz}$ is driven directly
by the vertical footfall force at the step rate, a straightforward
forced resonance, not the velocity-locked synchronous mechanism of
the Millennium Bridge. The Millennium effect is special to lateral
modes because pedestrians widen their stance to balance against
lateral deck motion, locking their lateral force to the deck velocity;
vertical gait does not synchronise to deck velocity the same way. The
diagnostic measurement is to record deck acceleration with crowd size
and pace: a direct forced resonance grows with the number of
in-step pedestrians roughly as $\sqrt{N}$ (random phase) up to the
fraction marching in step, whereas a synchronous instability shows a
sharp onset above a critical density. Confirm the $2\,\si{\hertz}$
vertical mode by an impact-hammer test with the bridge empty.
\end{exercise}

\begin{exercise}
A residential dishwasher in operation rattles the cabinet at a
particular speed during the spin cycle. Identify the most probable
dynamic mechanism (rotor imbalance, foundation resonance,
shipping-restraint pin not removed, or other) and the diagnostic
procedure.

\paragraph{Solution sketch.} A rattle that appears only at a
particular speed is a resonance: as the spin rate sweeps through a
band, its $1\times$ excitation crosses a natural frequency of the
cabinet, the mounting, or an adjacent panel, and the response peaks.
The most probable mechanism is a cabinet-panel or mounting resonance
excited by residual drum imbalance, sharpened if the unit is not level
or a packing restraint was left in. The diagnostic procedure: confirm
the rattle occurs only in a narrow speed band (resonance, not a loose
part that rattles at all speeds), then hold a hand on candidate panels
during the spin to find the one that goes quiet when damped. The fix
is to stiffen or damp the resonant panel, re-level the unit, or remove
the restraint, moving the natural frequency clear of the spin band.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Re-read the Federal Works Agency report on the Tacoma Narrows
Bridge collapse \cite{acc:tacoma-narrows-1941}. State the
structural-damping ratio of the deck's torsional mode estimated by
the report, the aerodynamic-damping term that turned the total
damping negative, and the critical wind speed at which the
transition occurred.

\paragraph{Solution sketch.} The exercise is a sourcing task: the
reader extracts the numbers from the report and the chapter's failure
section. The deck's torsional mode was very lightly damped,
$\zeta \approx 0.005$, typical of a welded-steel structure. The
aerodynamic-damping term arises from the pitching moment that depends
on angle of attack; above a critical wind speed of approximately
$19\,\si{\meter\per\second}$ it became negative and overwhelmed the
small structural damping, driving the total damping below zero and
making the torsional mode self-excited. The model answer states these
three figures and identifies the mechanism as negative aerodynamic
damping, not forced resonance at a fixed driving frequency.
\end{exercise}

\begin{exercise}
Re-read the Arup analysis of the Millennium Bridge lateral
oscillation \cite{acc:dallard-millennium-2001}. Identify the
critical pedestrian density, the lateral natural frequency of the
bridge, and the per-pedestrian lateral force coefficient that the
report uses to predict the onset of synchronous excitation.

\paragraph{Solution sketch.} A sourcing task. From the chapter's
failure section and the Arup analysis: the critical pedestrian density
was approximately $1.3\,\text{persons}/\si{\meter\squared}$; the
excited lateral modes lay near $1\,\si{\hertz}$, half the vertical
pacing rate; and the report models the per-pedestrian lateral force as
proportional to the deck's lateral velocity, $F = k v$, the
velocity-locked coupling that turns a positive feedback once the
coherent pedestrian forcing exceeds the structural damping. The model
answer reports the density, the frequency, and the velocity-proportional
force law, and notes that the random-phase $\sqrt{N}$ model fails once
synchronisation sets in.
\end{exercise}

\begin{exercise}
A high-speed turbine has lost a fan blade after a few hundred hours
of service. The blade's fracture surface shows fatigue striations.
State whether blade-pass resonance is a plausible cause, what
calculation would verify, and what design discipline (detuning,
damping, inspection interval) would have prevented the failure.

\paragraph{Solution sketch.} Fatigue striations indicate high-cycle
fatigue from a sustained alternating stress, exactly what a resonant
vibration produces. Blade-pass resonance is a plausible cause: if a
blade's natural frequency coincides with an excitation order (rotor
speed times number of upstream vanes, or an engine-order harmonic),
the blade vibrates at large amplitude and accumulates fatigue at its
hub fillet over hundreds of hours. The verifying calculation is a
Campbell diagram: plot blade natural frequencies against rotor speed,
overlay the engine-order excitation lines, and look for an
intersection inside the operating range. The preventive design
discipline is detuning the blade so no resonance falls in the
operating band, adding damping (friction dampers, coatings), and
setting an inspection interval below the predicted crack-growth life.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer argues that vibration analysis is overkill for a
small consumer product and that designing the product to be "stiff
enough" is sufficient. Write a one-paragraph reply that takes the
argument seriously and states when a vibration analysis is required.

\paragraph{Model-answer sketch.} A strong reply grants the kernel of
truth: stiffening raises natural frequencies, and pushing every
natural frequency above the highest excitation frequency does remove
resonance, so for a product with no internal sources and a benign
environment a stiff design can suffice. The reply then names the cases
where it does not: products with internal rotating parts (a fan, a
pump, a hard drive), products subjected to transport vibration or
drop, products with a resonance that stiffening cannot move out of
band without unacceptable mass, and products where a resonance
degrades function (a camera, a speaker, a sensor). In those cases a
vibration analysis is required, because "stiff enough" is an untested
claim until the natural frequencies and the excitation spectrum are
both known.
\end{exercise}

\begin{exercise}
A senior colleague proposes to mount a vibrating machine on
soft springs to reduce transmitted vibration. The machine's lowest
operating speed is $5\,\si{\hertz}$. Write a one-paragraph reply
that addresses the trade-off between steady-state isolation and
transient response, and states a defensible recommendation for the
isolator natural frequency.

\paragraph{Model-answer sketch.} Soft springs do isolate, provided the
isolator natural frequency sits well below the lowest operating
frequency so that $r > \sqrt2$ and the transmissibility drops below
one. With a lowest operating speed of $5\,\si{\hertz}$, a defensible
target is $f_n \approx 1.5$ to $2\,\si{\hertz}$ ($r \approx 2.5$ to
$3$), giving transmissibility around $0.1$ to $0.15$. The reply should
flag the cost of going softer: a low $f_n$ means large static
deflection, a flexible mount that sways under any transient (startup,
load change, a bump), and a slow-settling rocking response. The
recommendation balances the two: choose the highest $f_n$ that still
clears $\sqrt2$ with margin, add modest damping ($\zeta \approx 0.05$
to $0.1$) to survive startup through resonance, and verify the static
deflection is acceptable.
\end{exercise}

\begin{exercise}
Identify a vibrating system in your immediate environment whose
natural frequencies you have never measured (a desk, a window, a
ceiling fan, a bookshelf). State, in $200$ to $300$ words, the
mode shapes you would expect, the natural-frequency band, and the
mechanism by which the system might fail in service.

\paragraph{Model-answer sketch.} An open-ended reflection; a strong
answer names a specific object, sketches its lowest modes, and gives a
defensible frequency band with reasoning. For a bookshelf, the first
mode is a side-to-side sway of the whole unit on its base (mass of the
books, stiffness of the uprights in bending), likely a few hertz; the
second is a front-to-back rock; higher modes are shelf-plate bending,
tens of hertz. The failure mechanism is a resonance with a building
floor vibration or an earthquake that amplifies the sway until the
unit tips, the reason tall shelves are anchored to the wall. The
answer should connect the expected band to a plausible excitation and
name the resulting failure mode, demonstrating the frequency-coincidence
reasoning of the chapter.
\end{exercise}
