% Volume 8: Machines
% Opener: tagline + scope. Sits immediately after the \part{} declaration.

\vspace*{1.5cm}
\begin{center}
\itshape\large
Model becomes machine
\end{center}
\vspace{0.5cm}

\noindent
We turn the mathematics, physics, materials, and computing of Volumes I-VII into artifact: mechanism, manufacturing, electronics, sensors, robotics, mechatronic integration. The reader leaves able to design a machine end-to-end at an honest first-pass level: kinematic synthesis, stress and fatigue analysis, manufacturing-route selection, electronic interfacing, sensor integration, and control architecture.

\vspace{0.5cm}

\section*{What this volume is for}

Seven volumes have prepared the engineer to measure, model, force, energise, build out of matter, design with life, and reason about information. Volume VIII is where all of those threads meet a physical object the engineer is responsible for. A four-bar linkage is a mechanism with the kinematics of Volume III, the materials of Volume V, the manufacturing of this volume, the sensors and actuators of this volume, and a control problem that Volume IX will pick up. A motor drive is the electromagnetism of Volume IV, the materials of Volume V, the power electronics of this volume, and the embedded control of this volume. A robotic arm is all of the above plus the geometric reasoning of Volume II Chapter 9 and the algorithms of Volume VII Chapter 3. The volume is the synthesis point.

The standard of practice is what working engineers actually do. The reader does not leave able to design a turbine blade against a research-grade aerothermal model, or to derive a new motor topology from first principles. The reader leaves able to: choose a mechanism for a specified motion and verify that it works; select bearings, fasteners, and joints for a given load and lifetime; design a part for fatigue against an S-N curve and a stress concentration; produce a complete drawing pack a manufacturer will not reject; pick a manufacturing route for a part and write the tolerances that make the part interchangeable; design an analog signal-conditioning path; specify a power-conversion stage; build a working embedded system with the right sensors and the right communication bus; design a closed-loop position controller with the right motor and the right driver; reason about MEMS at the scaling level; build a small robot; integrate a mechatronic system with mechanical, electrical, and digital subsystems that do not fight each other.

\section*{What the reader brings}

Every chapter of this volume rests on the prior volumes. From Volume I: measurement discipline, error budgeting, calibration. From Volume II: vector calculus for forces and moments, ODEs for dynamics, linear algebra for kinematics, optimisation for design choices. From Volume III: statics, dynamics, vibrations, stress and strain. From Volume IV: thermodynamics for heat-of-friction and motor losses, electromagnetism for motors and generators, circuits for everything. From Volume V: materials selection, fatigue, fracture, corrosion. From Volume VI: not strictly required, but the biomechanics chapter has applications here. From Volume VII: algorithms, embedded computing, networks, control, signal processing.

The reader is also assumed to know what a working machine shop looks like, even if from a distance. The volume is unapologetic about presenting manufacturing tolerances, surface finishes, and the realities of part-to-print fidelity at engineering depth; a reader who has never seen a lathe or a CNC mill will find some passages abstract that working engineers find concrete.

\section*{The arc of the fourteen chapters}

The volume is organised in five arcs.

\textbf{Arc one: mechanism and structure} (Chapters 1-4). Linkages, cams, gears, the kinematic vocabulary. Bearings, lubrication, the friction discipline. Fasteners, springs, joints, the engineering of how parts connect. Fatigue and wear, the design under cyclic loading that turns Volume V's S-N curves into actual safe-life decisions.

\textbf{Arc two: representation and manufacture} (Chapters 5-7). CAD and engineering drawing, the documentation that makes parts manufacturable. Traditional manufacturing: machining, casting, forming. Additive manufacturing: the second-generation production method that has changed what is geometrically possible.

\textbf{Arc three: electrical} (Chapters 8-10). Analog electronics, the signal-conditioning and small-signal domain. Power electronics, the motor-drive and energy-conversion domain that dominates modern machines. Digital electronics and embedded systems, the computational substrate.

\textbf{Arc four: interfacing the machine to the world} (Chapters 11-12). Sensors and actuators at macro scale: the motors, encoders, force sensors that close the control loops. MEMS at micro scale: the same physics with stiction and packaging instead of bearings and bolts.

\textbf{Arc five: integration} (Chapters 13-14). Robotics: the integration of mechanism, sensing, computation, and control into machines that move under autonomous direction. Hydraulics, pneumatics, and the mechatronic integration that makes the whole machine work as a system.

\section*{What the reader will be able to do at the end}

A reader who finishes Volume VIII should be able to:

\begin{itemize}[leftmargin=*]
  \item Synthesise a planar linkage or cam mechanism for a stated motion
    profile and verify it analytically and in CAD.
  \item Select rolling-element bearings for a stated radial and axial
    load with an explicit $L_{10}$ life calculation.
  \item Specify a bolted, welded, or adhesive joint for a load and a
    fatigue life, including preload, friction coefficient, and
    inspection interval.
  \item Compute fatigue safe-life or damage-tolerant design margins
    against an S-N curve and a stress concentration; choose between
    the two design philosophies for a stated application.
  \item Produce a fully detailed engineering drawing of a part or
    assembly with GD\&T tolerances that a competent shop will not
    question.
  \item Choose a manufacturing route for a part (machined, cast,
    forged, formed, additively manufactured) and justify the choice
    against material, geometry, quantity, and tolerance.
  \item Design analog signal conditioning with op-amps and passive
    filters that meets a stated bandwidth, noise floor, and gain
    accuracy.
  \item Specify a power-conversion stage (buck, boost, inverter,
    motor drive) for a stated input-output relationship.
  \item Build a working embedded system on a microcontroller with the
    right sensors, the right buses, and the right firmware structure.
  \item Specify a sensor and actuator pair for a closed-loop control
    application and design the driver electronics.
  \item Reason about MEMS at the scaling level: which physics dominate
    at \si{\micro\meter} length scales and how packaging stress
    becomes the failure mode.
  \item Build a small mobile robot or robotic arm and have it
    accomplish a stated task.
  \item Integrate a mechatronic system of mechanical, electrical, and
    digital subsystems into one machine, document it, and hand it
    over for someone else to operate.
\end{itemize}

A reader who reaches Volume IX without these capabilities has read Volume VIII without doing it. We say so plainly because Volume IX assumes the substrate of this volume is well in hand: every control problem in Volume IX is a control problem on a machine of this volume's kind.

\section*{Archetypes introduced or developed in this volume}

Volumes I-VII introduced and developed: scaling, balance, failure, uncertainty, interface, transport, optimisation, stability, control, concurrency, ethics under capability. Volume VIII introduces no fresh archetypes; it deepens nearly all the existing ones in the substrate of the working machine.

\engterm{Scaling} returns in mechanism kinematics (Section 1.1), in bearing-life calculations (Section 2.5), in fatigue (Chapter 4), in MEMS (Chapter 12). The same physics behaves differently at different sizes: a bolt that holds an aircraft fuselage scales to a bolt that holds a wristwatch by changing both the geometry and the dominant failure mode.

\engterm{Failure} returns in the failure section of every chapter, and in the dedicated Chapter 4 on fatigue-and-wear design. The failure modes are specific and physical: gear-tooth pitting, bearing spalling, bolt-thread stripping, fatigue crack initiation at a stress concentration, weld undercut, casting porosity, additive-manufactured anisotropy, capacitor end-of-life, MEMS stiction, joint creep. Each is a name for a way a machine stops doing what it was designed for.

\engterm{Interface} returns at every mechanical-electrical-digital boundary. The mechatronic integration of Chapter 14 is the volume's culminating engagement with this archetype.

\engterm{Control} returns in mechanism design (selecting a kinematic ratio to control motion), in motor drives (Chapter 9), in embedded firmware (Chapter 10), in actuator electronics (Chapter 11), in robotics (Chapter 13). The full control-theoretic treatment lands in Volume IX.

\engterm{Optimisation} returns in mechanism synthesis, in topology-optimised additive parts, in motor-drive efficiency, in manufacturing process selection.

\section*{The volume project}

The reader designs, builds, and characterises a complete machine at small scale. Examples: a 3-axis CNC, a small robotic arm, a precision linear stage, a bench-top centrifuge, a CV-based sorter, a small autonomous vehicle. The deliverable is a full engineering pack: requirements, drawings, BOM (bill of materials), FMEA (failure modes and effects analysis), controller design, software, test data, error budget, manual. The project is the volume's deliverable; the engineering pack is the credential by which the reader earns admission to Volume IX.

The project is unapologetically ambitious. A working four-axis robot of the kind a hobbyist can build in a basement is approximately the right scale; a working CNC mill of the kind a small machine shop can use is the upper end. The reader who chooses smaller (a single-axis stage with a working encoder) does the same disciplines on a smaller scale; the deliverable is the same set of documents.

\section*{A note on the next volume}

Volume IX takes up Systems. Every machine of this volume is a system already (Chapter 14 makes the point explicit). Many machines, many control loops, many subsystems working together without lying to themselves are the next problem. The bridge from this volume is direct: every working closed-loop controller in this volume is a Volume IX entry-point, every sensor-fusion gesture is a Volume IX foundation, every system-of-systems mechatronic integration is a Volume IX practice in miniature.

\vspace{1cm}
