\chapter{Acoustics: waves in solids and fluids}
\label{vol03:ch07}

\epigraph{The sensation of a musical tone is due to a rapid periodic
motion of sonorous bodies; that of a noise to non-periodic motions.}{
Hermann von Helmholtz, \emph{On the Sensations of Tone}, English
edition of 1875, translated and annotated by Alexander J.\ Ellis from
the German \emph{Die Lehre von den Tonempfindungen} of 1863
\cite[p.~8]{hist:helmholtz1863}.}

\chapmeta{Half-life: durable. Archetypes: transport, scaling, stability. Exercise target: 30. Project track: physical measurement plus analysis (hybrid). Hazard class: low (speaker measurements with hearing-protection at high levels). Reader path default: standard, with sections explicitly tagged. Named cases: none.}

\noindent
A pressure disturbance in a fluid or an elastic disturbance in a
solid propagates as a wave. The wave carries energy, transmits
information, and at high enough amplitude does mechanical work on
the boundaries it reaches. The engineering of acoustics is the
engineering of those waves: how they are generated, how they
propagate, how they reflect from interfaces, how they couple to
structures that they cause to vibrate, and how they are measured or
controlled. This chapter follows the wave through its domains:
acoustic waves in fluids, elastic waves in solids, the interfaces
between them, the engineered environments (rooms, oceans, the
atmosphere) in which waves live, the engineering uses of ultrasonic
and infrasonic frequencies, and the structural failures that
high-intensity acoustic loading can cause.

\section{The wave equation in fluids}\pathtag{core}

A small pressure perturbation $p(\vec{x}, t)$ in a fluid otherwise at
rest, with density perturbation $\rho'(\vec{x}, t)$ and velocity
perturbation $\vec{u}(\vec{x}, t)$, satisfies three coupled
linearised equations: conservation of mass, conservation of momentum
(Euler's equation, ignoring viscosity), and a thermodynamic relation
connecting the pressure to the density. We derive the wave equation
from those three in one dimension, because the algebra is short and the
three physical inputs stay visible.

Start from the exact one-dimensional conservation laws for an inviscid
fluid. Mass conservation (the continuity equation) is
\[
\frac{\partial \rho}{\partial t}
+ \frac{\partial (\rho u)}{\partial x} = 0,
\]
and momentum conservation (the Euler equation, no viscosity, no body
force) is
\[
\rho\left(\frac{\partial u}{\partial t}
+ u\frac{\partial u}{\partial x}\right)
= -\frac{\partial p}{\partial x}.
\]
Now write each field as a steady ambient value plus a small
perturbation: $\rho = \rho_0 + \rho'$, $p = p_0 + p'$, and a velocity
$u$ that is itself small (the ambient fluid is at rest, $u_0 = 0$). The
ambient state is uniform and static, so $\rho_0$ and $p_0$ are constants
in space and time. Substitute and keep only first-order terms,
discarding every product of two small quantities. The convective term
$u\,\partial_x u$ is second order and drops; so is $\rho' \partial_t u$
beside $\rho_0 \partial_t u$. The linearised pair is
\[
\frac{\partial \rho'}{\partial t}
+ \rho_0\frac{\partial u}{\partial x} = 0,
\qquad
\rho_0\frac{\partial u}{\partial t}
= -\frac{\partial p'}{\partial x}.
\]
The third input closes the system: the pressure responds to the density
through the fluid's compressibility. For a small isentropic change,
\[
p' = \left(\frac{\partial p}{\partial \rho}\right)_s \rho'
   \equiv c^2\,\rho',
\qquad
c^2 \equiv \left(\frac{\partial p}{\partial \rho}\right)_s,
\]
which defines the constant $c$. The derivative is taken at constant
entropy because the acoustic compression and rarefaction are fast
compared to thermal-equilibration times in ordinary gases and liquids;
the process is isentropic to high accuracy.

Eliminate the velocity. Differentiate the linearised continuity
equation with respect to time and the linearised momentum equation with
respect to space:
\[
\frac{\partial^2 \rho'}{\partial t^2}
+ \rho_0\frac{\partial^2 u}{\partial x\,\partial t} = 0,
\qquad
\rho_0\frac{\partial^2 u}{\partial t\,\partial x}
= -\frac{\partial^2 p'}{\partial x^2}.
\]
The mixed velocity derivative is common to both (the order of
differentiation does not matter for a smooth field), so subtracting
removes it:
\[
\frac{\partial^2 \rho'}{\partial t^2}
= \frac{\partial^2 p'}{\partial x^2}.
\]
Replace $\rho' = p'/c^2$ from the state relation, and the
\engterm{acoustic wave equation} for the pressure perturbation follows,
written here in its three-dimensional form, to which the same
elimination leads from the vector continuity and Euler equations:
\[
\frac{\partial^2 p}{\partial t^2} = c^2 \nabla^2 p,
\]
with the wave speed
\[
c = \sqrt{\left(\frac{\partial p}{\partial \rho}\right)_s} =
\sqrt{\frac{K_s}{\rho_0}},
\]
where $K_s = \rho_0 (\partial p/\partial\rho)_s$ is the adiabatic bulk
modulus of the fluid and $\rho_0$ its undisturbed density. The two
expressions for $c$ are the same relation, since the bulk modulus is by
definition $K_s = \rho_0 (\partial p/\partial\rho)_s$. The derivation
fixes three assumptions for everything that follows: small amplitude
(every retained term is first order in the perturbation), no viscosity
(the Euler equation, so no acoustic absorption appears here; section~7.4
restores it), and isentropic compression (the stiffness is the adiabatic
modulus, not the isothermal one, the correction Newton missed and Laplace
supplied).

For an ideal gas the adiabatic bulk modulus is $K_s = \gamma p_0$,
with $\gamma = c_p/c_v$ the ratio of specific heats, giving
\[
c = \sqrt{\frac{\gamma p_0}{\rho_0}} = \sqrt{\gamma R T_0},
\]
where $R = R_{\text{universal}}/M$ is the specific gas constant for
the gas and $T_0$ the absolute temperature. The speed of sound in
air at $20^{\circ}\,\mathrm{C}$ and atmospheric pressure is
approximately $343\,\si{\meter\per\second}$; the dependence on
temperature is $c \propto \sqrt{T}$, so a $1^{\circ}\,\mathrm{C}$
change shifts the speed by about $0.6\,\si{\meter\per\second}$. The
pressure dependence at fixed temperature is zero for an ideal gas;
real gases show small pressure corrections.

In water, $c \approx 1{,}480\,\si{\meter\per\second}$ at $20^{\circ}\,
\mathrm{C}$, weakly dependent on temperature, salinity, and pressure;
the standard oceanographic formulas (Chen-Millero, UNESCO 1983) give
$c$ to within $\sim 0.1\,\si{\meter\per\second}$ across the
temperature-salinity-depth range of the world's oceans. In steel,
the longitudinal wave speed is $\sim 5{,}900\,\si{\meter\per\second}$;
in aluminum, $\sim 6{,}300\,\si{\meter\per\second}$. The pattern is
that solids are faster than dense liquids, dense liquids are faster
than thin gases, and the differences span an order of magnitude.

Plane-wave solutions to the acoustic wave equation are
\[
p(x, t) = A\,e^{i(kx - \omega t)} + \text{c.c.},
\]
with the dispersion relation $\omega = c k$, so the wavelength
$\lambda = 2\pi/k$ and the frequency $f = \omega/(2\pi)$ satisfy
$\lambda = c/f$. Substituting the trial form into the wave equation
returns $-\omega^2 = c^2(-k^2)$, which is the dispersion relation, and
its linearity in $\omega$ versus $k$ is what makes acoustic propagation
non-dispersive: all audible frequencies travel at the same speed $c$, so
a complex waveform keeps its shape as it propagates, the reason a spoken
sentence arrives intelligible across a room. Audible frequencies range
from $\sim 20\,\si{\hertz}$ to $\sim 20\,\si{\kilo\hertz}$,
corresponding to wavelengths in air from $\sim 17\,\si{\meter}$ to
$\sim 17\,\si{\milli\meter}$, a span of three orders of magnitude. The
wavelength sets the scale at which the wave's geometry matters: a wall
larger than $\lambda$ reflects a wave cleanly; a wall smaller than
$\lambda$ is acoustically transparent.

The plane wave carries a fixed relation between its pressure and the
particle velocity it drives. Insert the travelling pressure $p =
A\cos(kx-\omega t)$ into the linearised momentum equation $\rho_0
\partial_t u = -\partial_x p$. The space derivative is
$\partial_x p = -kA\sin(kx-\omega t)$, so $\rho_0 \partial_t u =
kA\sin(kx-\omega t)$, and integrating in time gives
\[
u = \frac{k}{\rho_0 \omega}A\cos(kx-\omega t) = \frac{p}{\rho_0 c},
\]
using $\omega/k = c$. Pressure and particle velocity are in phase and
proportional for a forward-running plane wave, and their ratio is the
\engterm{characteristic acoustic impedance}
\[
Z_0 = \frac{p}{u} = \rho_0 c,
\]
the acoustic analogue of the resistance that relates voltage to current
in a circuit. A wave running in the $-x$ direction carries the opposite
sign, $u = -p/(\rho_0 c)$, the fact the reflection-coefficient
derivation of section~7.3 turns on.

\engterm{Acoustic intensity} is the rate of energy transport per unit
area, in $\si{\watt\per\meter\squared}$, and it follows from the
pressure-velocity relation. Intensity is the time average of the
instantaneous power flux $p\,u$ through unit area,
\[
I = \langle p\,u\rangle
  = \left\langle \frac{p^2}{\rho_0 c}\right\rangle
  = \frac{\langle p^2\rangle}{\rho_0 c}
  = \frac{p_{\text{rms}}^2}{\rho_0 c},
\]
where the second step uses $u = p/(\rho_0 c)$ and the last uses the
definition of the root-mean-square pressure
$p_{\text{rms}}^2 = \langle p^2\rangle$. For a single tone of amplitude
$A$ the time average of the square is $A^2/2$, so $p_{\text{rms}} =
A/\sqrt 2$, the same factor familiar from alternating-current power. Air
at sea level has $\rho_0 c \approx 415\,\si{\kilogram\per\meter\squared
\per\second}$ ($\si{\pascal\second\per\meter}$, the rayl); water has
$\rho_0 c \approx 1.5 \times 10^{6}\,\text{rayl}$, almost four orders
of magnitude larger. The impedance mismatch between air and water is
why a clap above the water is barely audible below, and why a
submarine's sonar emitter is not a loudspeaker.

A source of finite size radiates spherical waves, not plane waves, and
the geometry changes the decay law. A spherically symmetric solution of
the wave equation has the form
\[
p(r, t) = \frac{A}{r}\,e^{i(kr - \omega t)},
\]
which the substitution $p = \psi(r,t)/r$ converts into the
one-dimensional wave equation for $\psi$, so the radial wave propagates
at the same speed $c$ but its amplitude falls as $1/r$. The intensity,
quadratic in pressure, therefore falls as $1/r^2$, which is energy
conservation made visible: the same acoustic power $W$ crosses every
sphere of area $4\pi r^2$, so $I = W/(4\pi r^2)$ and the level drops six
decibels per doubling of distance, the point-source slope of
figure~\ref{fig:vol03ch07:spherical-spreading}. Close to the source,
within a wavelength or so, the pressure and velocity fall out of phase
(the near field carries reactive energy that sloshes back and forth
without net transport); far from the source the spherical wave looks
locally plane, pressure and velocity return to phase, and the
plane-wave impedance $\rho_0 c$ and intensity relation recover. The
crossover at $kr\sim 1$ separates the near field, where a microphone
reads a pressure that does not represent radiated power, from the far
field, where it does, the distinction the sound-power survey later in the
chapter depends on.

\input{volumes/03-force/ch07-acoustics/figures/fig-spherical-spreading.tex}

Three logarithmic scales recur, and they differ by what sits in the
reference and by the factor in front. A \engterm{decibel} is ten times
the base-ten logarithm of a power ratio,
\[
L = 10\log_{10}\frac{W}{W_{\text{ref}}}\,\si{\decibel},
\]
so a factor-of-ten change in power is ten decibels and a doubling is
$10\log_{10}2 \approx 3\,\si{\decibel}$. The factor becomes twenty
whenever the quantity in the ratio is an amplitude rather than a power,
because power goes as amplitude squared and $\log(x^2) = 2\log x$. Sound
intensity and sound power are power quantities and carry the factor ten;
sound pressure is an amplitude and carries the factor twenty. The
\engterm{sound-power level} of a source is
$L_W = 10\log_{10}(W/W_0)$ with $W_0 = 10^{-12}\,\si{\watt}$, a property
of the source alone. The \engterm{sound-intensity level} is
$L_I = 10\log_{10}(I/I_0)$ with $I_0 = 10^{-12}\,\si{\watt\per
\meter\squared}$, a property of the field at a point.

\engterm{Sound pressure level} (SPL) is the logarithmic measure
\[
\text{SPL} = 20\log_{10}\left(\frac{p_{\text{rms}}}{p_{\text{ref}}}
\right)\,\si{\decibel},
\]
with the reference pressure $p_{\text{ref}} = 20\,\si{\micro\pascal}$
in air, conventionally close to the threshold of hearing at $1\,\si{
\kilo\hertz}$. The three scales nearly coincide by the deliberate choice
of references: in a free plane-wave field $I = p_{\text{rms}}^2/(\rho_0
c)$, and since $\rho_0 c \approx 400\,\text{rayl}$ in air makes
$I_0 \rho_0 c \approx p_{\text{ref}}^2$ within a fraction of a decibel,
the intensity level and the sound pressure level read the same number in
the far field. The coincidence is a convenience of the reference values,
not an identity, and it fails near a source or in a reverberant field
where pressure and velocity are not in phase. Normal conversation is at about $60\,\si{\decibel}$
SPL; a busy urban street, $80\,\si{\decibel}$; a power-tool at one
metre, $100\,\si{\decibel}$; a jet engine at $30\,\si{\meter}$,
$130\,\si{\decibel}$; the pain threshold is around $140\,\si{
\decibel}$. The decibel scale's logarithmic nature reflects both the
range of intensities the human ear handles ($\sim 12$ orders of
magnitude in intensity, $\sim 6$ in pressure) and the psychophysical
near-logarithmic response of the auditory system.

Two arithmetic habits make the decibel scale usable at sight. First,
a doubling of pressure is $20\log_{10} 2 \approx 6\,\si{\decibel}$, and
a doubling of intensity (or acoustic power) is $10\log_{10} 2 \approx
3\,\si{\decibel}$. Second, levels from incoherent sources combine on an
energy basis, not by adding decibels: two equal incoherent sources sum
to $+3\,\si{\decibel}$, ten equal sources to $+10\,\si{\decibel}$. The
general rule for combining band levels $L_i$ is
\[
L_{\text{total}} = 10\log_{10}\!\left(\sum_i 10^{L_i/10}\right)\,
\si{\decibel},
\]
the same energy sum that \texttt{code/spl\_octave.py} performs across
octave bands. A consequence worth internalising: adding a quiet source
to a loud one barely moves the total, because the energy sum is
dominated by its largest term. A $60\,\si{\decibel}$ source added to an
$80\,\si{\decibel}$ source raises the total by $0.04\,\si{\decibel}$,
which is to say not at all. Noise control that ignores the loudest path
is wasted effort.

\begin{archetype}[Transport]
Acoustics is the transport archetype carried by a wave rather than by a
flow. The conserved currency is energy; the carrier is a pressure
disturbance; the medium's resistance to carrying it is the
characteristic impedance $\rho_0 c$. Every later section is a
specialisation of one transport question. Reflection and transmission
ask how much energy crosses a boundary where the impedance changes.
Atmospheric and underwater attenuation ask how the energy current
decays along a path. Room acoustics asks where the transported energy
piles up into standing patterns. Noise control asks how to interrupt
the transport between a source and a receiver. Holding the transport
frame in view keeps the chapter's many formulas anchored to one
accounting question: where does the acoustic energy go, and how much
of it arrives.
\end{archetype}

\begin{mastery}
\textbf{The velocity potential and the radiation integral.} The pressure
wave equation hides a tidier object underneath it. Because the linearised
flow is irrotational (the curl of the linearised Euler equation gives
$\partial_t(\nabla\times\vec u) = 0$, so an initially irrotational field
stays so), the velocity derives from a scalar \engterm{velocity potential}
$\phi$ through $\vec u = \nabla\phi$, and the pressure follows from
$p = -\rho_0\,\partial_t\phi$. Both $\phi$ and $p$ satisfy the same wave
equation, but $\phi$ is the natural unknown for radiation problems because
boundary motion enters through its gradient. A source that adds volume to
the medium, a pulsating sphere or a vibrating piston, drives the
\engterm{inhomogeneous} wave equation
\[
\nabla^2 \phi - \frac{1}{c^2}\frac{\partial^2\phi}{\partial t^2}
= -q(\vec x, t),
\]
with $q$ the volume-source density. The solution is the retarded-potential
integral, the acoustic form of the Green's-function solution that
electromagnetism shares,
\[
\phi(\vec x, t) = \frac{1}{4\pi}\int
\frac{q\!\left(\vec x', \, t - |\vec x - \vec x'|/c\right)}
     {|\vec x - \vec x'|}\,\dd V',
\]
which says the potential at a point now is the sum of contributions from
every source element, each weighted by $1/r$ for spreading and delayed by
the travel time $r/c$. A compact source of fluctuating volume velocity $Q$
reduces this to the \engterm{monopole} field $p = (i\omega\rho_0 Q/4\pi r)
e^{i(kr-\omega t)}$, whose $1/r$ amplitude and radiated power $W =
\rho_0 c k^2 |Q|^2/(8\pi)$ underlie the loudspeaker-radiator worked example
and the sound-power survey. Two adjacent monopoles in antiphase make a
\engterm{dipole}, the model of an unbaffled cone or a small vibrating
rigid body, whose radiation falls off faster with angle and far more
steeply with frequency ($W\propto\omega^4$), the reason a bare driver
makes no bass and a baffle exists to suppress the rear wave. The radiation
integral is how every source in the chapter, from a transducer face to a
machine surface, is turned from a prescribed motion into a radiated field.
\end{mastery}

\section{The wave equation in solids}\pathtag{core}

A solid supports two distinct kinds of mechanical waves, in contrast
to a fluid's single longitudinal acoustic mode. The two modes
correspond to the solid's independent elastic constants from
chapter 3.

\engterm{Longitudinal (P) waves}: the particle motion is along the
direction of wave propagation, as in a fluid. The wave speed is
\[
c_L = \sqrt{\frac{\lambda + 2\mu}{\rho}} = \sqrt{\frac{E(1-\nu)}{
\rho(1+\nu)(1-2\nu)}},
\]
with $\lambda, \mu$ the Lam\'e constants of chapter 3, $E$ Young's
modulus, $\nu$ Poisson's ratio. In structural steel,
$c_L \approx 5{,}900\,\si{\meter\per\second}$; in aluminum,
$\sim 6{,}300\,\si{\meter\per\second}$; in granite, $\sim 5{,}500\,\si
{\meter\per\second}$; in concrete, $\sim 3{,}500\,\si{\meter\per\second}$.
Reference densities, speeds, and impedances for these media and others
are tabulated in \texttt{data/sound\_speed\_materials.csv}.

\engterm{Transverse (S) waves}: the particle motion is perpendicular
to the propagation direction. Transverse waves exist in solids
because the material resists shear deformation; fluids have no
static shear resistance and therefore no transverse waves at low
frequency. The transverse wave speed is
\[
c_T = \sqrt{\frac{\mu}{\rho}} = \sqrt{\frac{G}{\rho}} = \sqrt{
\frac{E}{2\rho(1+\nu)}}.
\]
In structural steel, $c_T \approx 3{,}200\,\si{\meter\per\second}$,
about $0.55\,c_L$. The ratio $c_T/c_L$ is determined by Poisson's
ratio; for typical engineering materials with $\nu \approx 0.3$, the
ratio is $\sim 0.5$.

In an unbounded solid, both wave types coexist; an arbitrary
disturbance decomposes into longitudinal and transverse parts that
propagate at their respective speeds. At a free surface or at an
interface with a different medium, the two wave types couple: an
incident longitudinal wave generates both reflected and refracted
longitudinal and transverse waves (this is mode conversion). The
boundary-condition algebra is the content of section 7.3.

The wave equations for the two modes have the standard scalar form
$\partial^2 \phi/\partial t^2 = c^2 \nabla^2 \phi$ with the
respective wave speeds. For applications in earthquake seismology,
ultrasonic testing of materials, and underground propagation, both
wave types must be analysed; in pure-fluid problems (audio, sonar,
atmospheric acoustics) only the longitudinal mode appears.

\engterm{Surface (Rayleigh) waves}: at the free surface of a solid,
a third wave type exists, the Rayleigh wave, in which particle
motion is elliptical in a plane containing the propagation direction
and the surface normal, with amplitude decaying exponentially with
depth. Rayleigh waves were discovered by Lord Rayleigh in 1885 as a
consequence of his \emph{Theory of Sound} analysis
\cite{hist:rayleigh1877}. They are the dominant carrier of
high-frequency surface vibration and are exploited in surface-
acoustic-wave (SAW) devices in modern electronics and in non-
destructive ultrasonic surface-flaw inspection (chapter 12 of Volume
X). The Rayleigh wave speed is between $c_T$ and a fraction (about
$0.9$) of $c_T$, depending on Poisson's ratio.

\engterm{Guided (Lamb) waves}: when the solid is a plate rather than a
half-space, the two free surfaces confine the wave and the longitudinal and
shear modes couple into a family of guided \engterm{Lamb waves}. Each mode
has a symmetric or antisymmetric displacement pattern about the plate
midplane, and each is dispersive: its speed depends on the product of
frequency and plate thickness, so a pulse spreads as it travels, unlike the
non-dispersive bulk waves. The lowest antisymmetric mode is a bending wave
that at low frequency-thickness product moves far slower than the shear
speed and carries the flexural vibration of thin panels, the mode the
coincidence-frequency and mass-law calculations of the noise-control
sections turn on. The lowest symmetric mode is an extensional wave close to
the plate longitudinal speed. Guided waves are what ultrasonic inspection
actually uses on pipes, aircraft skins, and rail: a single transducer
launches a Lamb mode that runs metres along the structure and returns
echoes from corrosion, cracks, and disbonds far from the probe, so the
inspection covers area rather than a single thickness line. Selecting a mode
and a frequency-thickness point on the dispersion curve is the design choice
that trades penetration, sensitivity to a given defect type, and the
distance the mode travels before dispersion smears it beyond use. The two
lowest modes carry the phase-velocity dispersion of
figure~\ref{fig:vol03ch07:lamb-dispersion}: the antisymmetric $A_0$
flexural mode climbs from far below the shear speed toward the surface-wave
speed as the wavelength drops below the plate thickness, while the
symmetric $S_0$ extensional mode falls from the plate longitudinal speed
toward the shear speed.

\input{volumes/03-force/ch07-acoustics/figures/fig-lamb-dispersion.tex}

Mode conversion at an interface is not a nuisance but a design lever. When a
longitudinal wave in a coupling wedge strikes a solid surface obliquely, the
refracted longitudinal and shear waves leave at different angles set by
Snell's law with their different speeds, and between the two longitudinal
and shear critical angles only the shear wave propagates into the solid. A
probe wedge angled into that window launches a pure shear or a pure surface
wave into the part, the standard angle-beam technique for finding cracks
that a normal-incidence beam would miss because they lie perpendicular to
the surface. The same mode conversion complicates seismology, where a P-wave
striking a layer boundary spawns reflected and transmitted P- and S-waves
whose separate arrivals a seismogram must disentangle, and it is why the
fluid-solid boundary-condition system of the next section carries three
transmitted amplitudes rather than one.

\section{Reflection, refraction, transmission at interfaces}
\pathtag{core}

A plane wave incident on an interface between two media of
characteristic impedances $Z_1 = \rho_1 c_1$ and $Z_2 = \rho_2 c_2$
is partly reflected and partly transmitted. For normal incidence on
a fluid-fluid interface, the pressure-amplitude reflection coefficient
is
\[
R = \frac{Z_2 - Z_1}{Z_2 + Z_1},
\]
and the transmission coefficient is $T = 1 + R = 2Z_2/(Z_1 + Z_2)$.
The intensity reflection coefficient is $|R|^2$; the intensity
transmission is $1 - |R|^2$.

For an air-water interface ($Z_{\text{air}} \approx 415$, $Z_{\text{
water}} \approx 1.5\times 10^{6}\,\text{rayl}$), the pressure
reflection coefficient is approximately $R \approx 0.9994$, so the
intensity reflected from above is $\sim 99.9\,\%$ and the intensity
transmitted into water is $\sim 0.1\,\%$. The impedance mismatch
makes air and water nearly acoustically isolated, which is why
submarines can hide from airborne sound and why a swimmer underwater
hears only muffled fragments of conversation above the surface.

\input{volumes/03-force/ch07-acoustics/figures/fig-impedance-reflection.tex}

The intensity transmission coefficient written in terms of the
impedance ratio $r = Z_2/Z_1$ is
\[
T_I = 1 - |R|^2 = \frac{4 Z_1 Z_2}{(Z_1 + Z_2)^2}
    = \frac{4r}{(1+r)^2},
\]
plotted in figure~\ref{fig:vol03ch07:impedance-reflection}. The form is
symmetric in $r$ and $1/r$: a tenfold mismatch transmits the same
fraction in either direction. The cure for a large mismatch is to
bridge it in stages. A quarter-wavelength layer of intermediate
impedance $Z_m = \sqrt{Z_1 Z_2}$ makes the interface perfectly
transmitting at the design frequency, the acoustic version of the
optical anti-reflection coating.

The matching condition is worth deriving, because it shows why the
intermediate impedance is the geometric mean and why the layer must be a
quarter wavelength. A finite layer of impedance $Z_m$ and thickness $d$
between medium 1 and a load medium of impedance $Z_2$ transforms the load
impedance into an input impedance seen from medium 1. The transmission-line
transformation, derived in full as the transfer-matrix box later in the
chapter, is
\[
Z_{\text{in}} = Z_m\,
\frac{Z_2 + i Z_m\tan(k_m d)}{Z_m + i Z_2\tan(k_m d)},
\]
with $k_m$ the wavenumber in the layer. Set the layer a quarter
wavelength thick, $k_m d = \pi/2$, so $\tan(k_m d)\to\infty$ and the
finite terms drop beside the $\tan$ terms:
\[
Z_{\text{in}} = Z_m\,\frac{Z_m}{Z_2} = \frac{Z_m^2}{Z_2}.
\]
The interface is reflection-free when the input impedance equals the
source impedance, $Z_{\text{in}} = Z_1$, which requires
$Z_m^2/Z_2 = Z_1$, that is $Z_m = \sqrt{Z_1 Z_2}$. The quarter-wave layer
inverts the load about $Z_m$, and choosing $Z_m$ as the geometric mean
places $Z_1$ and $Z_2$ symmetrically on either side of the inversion, so
the source sees its own impedance and nothing reflects. The match is exact
only at the design frequency, where the thickness is exactly a quarter
wavelength; away from it the $\tan$ term departs from infinity and the
reflection grows, which is why a single layer matches over a moderate
band and broadband transducers stack several graded layers. Medical
ultrasound gel, the matching layer on a transducer face, and the
impedance-graded horn of a loudspeaker all exploit this. A worked
instance closes the chapter.

For oblique incidence at angle $\theta_1$ from the surface normal,
Snell's law gives the refracted angle $\theta_2$:
\[
\frac{\sin\theta_1}{c_1} = \frac{\sin\theta_2}{c_2}.
\]
The form is identical to Snell's law in optics; the wave-mechanical
content is the same. If $c_2 > c_1$, the refracted wave bends away
from the normal, and at a critical incident angle
$\theta_c = \arcsin(c_1/c_2)$ the refracted wave runs along the
interface (total internal reflection for incident waves at greater
angles). The total-internal-reflection regime is exploited in
acoustic waveguides (sound channels in the ocean, the SOFAR channel)
and in fibre-optic-like acoustic delay lines.

At a fluid-solid interface, the analysis is richer because the solid
supports both longitudinal and transverse modes. An incident fluid
acoustic wave generates a reflected fluid acoustic wave plus a
refracted longitudinal wave and a refracted transverse wave in the
solid. The boundary conditions (continuity of normal stress, zero
shear stress at the fluid-solid interface for a frictionless interface,
matching normal particle displacement) yield a $3 \times 3$ algebraic
system for the three transmitted wave amplitudes; the algebra is
standard and is in the standard acoustics texts \cite[chs.~6,
8]{text:kinsler-acoustics}.

The geometry of that system is more transportable than its algebra. Snell's
law holds separately for each refracted mode, so the incident angle
$\theta_1$ in the fluid maps to a longitudinal refraction angle $\theta_L$
through $\sin\theta_1/c_1 = \sin\theta_L/c_L$ and to a shear refraction
angle $\theta_T$ through $\sin\theta_1/c_1 = \sin\theta_T/c_T$. Because the
solid's longitudinal speed exceeds its shear speed, the two refracted rays
leave at different angles, and each has its own critical angle where the
refracted ray runs along the interface and the transmitted mode goes
evanescent. Water on steel makes the arithmetic concrete. With
$c_1 = 1480\,\si{\meter\per\second}$, $c_L = 5900\,\si{\meter\per\second}$,
and $c_T = 3200\,\si{\meter\per\second}$, the first (longitudinal) critical
angle is $\theta_{cL} = \arcsin(1480/5900) \approx 14.5^{\circ}$ and the
second (shear) critical angle is $\theta_{cT} = \arcsin(1480/3200) \approx
27.6^{\circ}$. Below $14.5^{\circ}$ both modes enter the solid; between
$14.5^{\circ}$ and $27.6^{\circ}$ only the shear wave propagates in, since
the longitudinal mode has already gone evanescent; above $27.6^{\circ}$ both
modes are evanescent and the interface totally reflects. The window between
the two critical angles is the working principle of the angle-beam
ultrasonic probe: a coupling wedge cut to place the incident angle in that
band launches a pure shear wave into the part and nothing else, so the
inspector reads a clean shear-wave field with no longitudinal ghost. The
same two-critical-angle structure governs a seismic wave crossing a
sediment-to-rock boundary and the leaky-Rayleigh wave a fluid loading puts
on a solid surface.

\section{Sound propagation in air, water, ground}\pathtag{standard}

The acoustic wave equation describes free-field propagation. Real
environments add attenuation, refraction by gradients in the medium,
scattering by boundaries, and ground or surface reflections that
shape the received signal at any point.

\subsection{Atmospheric propagation}

Sound in the atmosphere is attenuated by three mechanisms:
geometric spreading (a $1/r$ decay for a spherical wave's pressure
amplitude, $1/r^2$ for intensity), classical absorption (viscous and
thermal-conduction losses in the air, proportional to $\omega^2$ at
low frequencies), and molecular-relaxation absorption (energy
transfer to and from rotational and vibrational modes of N$_2$ and
O$_2$, large in the kHz band and humidity-dependent). The total
atmospheric attenuation in $\si{\decibel\per\kilo\meter}$ varies
strongly with frequency and humidity; at $1\,\si{\kilo\hertz}$ in
dry air it is $\sim 5\,\si{\decibel\per\kilo\meter}$, at $10\,\si{
\kilo\hertz}$ it is $\sim 100\,\si{\decibel\per\kilo\meter}$, at
$100\,\si{\hertz}$ it is $\sim 0.5\,\si{\decibel\per\kilo\meter}$.
Low frequencies travel far; high frequencies do not.

\input{volumes/03-force/ch07-acoustics/figures/fig-atmospheric-attenuation.tex}

Figure~\ref{fig:vol03ch07:atmospheric-attenuation} shows the frequency
dependence at two humidities; the reference numbers are tabulated in
\texttt{data/atmospheric\_absorption.csv}. The molecular-relaxation
contribution is what makes the curves cross rather than scale
uniformly with humidity, so a rule of thumb stated for one humidity
fails at another. The single transportable fact is the steep climb
with frequency: distant sources lose their high-frequency content
first, which is why far thunder rumbles and near thunder cracks.

The classical part of that attenuation has a closed form worth carrying,
because it fixes the floor beneath the humidity-dependent excess. Viscosity
and heat conduction together give the Stokes-Kirchhoff absorption
coefficient
\[
\alpha_{\mathrm{cl}} = \frac{\omega^2}{2\rho_0 c^3}
  \left[\frac{4}{3}\mu + \mu_B + \kappa\left(\frac{1}{c_v}-\frac{1}{c_p}
  \right)\right],
\]
with $\mu$ the shear viscosity, $\mu_B$ the bulk viscosity, $\kappa$ the
thermal conductivity, and $c_v,c_p$ the specific heats. The bracket is a
fixed property of the gas, so the frequency dependence is the leading
$\omega^2$, the straight line of gradient two in
figure~\ref{fig:vol03ch07:classical-absorption}. For air the whole bracket
collects into a convenient engineering constant: the classical coefficient
is about $1.4\times 10^{-7}\,f^2$ in decibels per kilometre with $f$ in
hertz, so $1\,\si{\kilo\hertz}$ loses about $0.14\,\si{\decibel\per\kilo
\meter}$ to the classical mechanism, $10\,\si{\kilo\hertz}$ about
$14\,\si{\decibel\per\kilo\meter}$, and a $1\,\si{\mega\hertz}$ airborne
ultrasonic beam about $1.4\times 10^{5}\,\si{\decibel\per\kilo\meter}$, or
$140\,\si{\decibel\per\meter}$, which is why airborne ultrasound reaches
only centimetres. The measured atmospheric loss of the earlier figure sits
above this line through the audible band because molecular relaxation adds a
humidity-dependent excess; the classical curve is the irreducible remainder
that survives when the relaxation contribution is subtracted.

\input{volumes/03-force/ch07-acoustics/figures/fig-classical-absorption.tex}

Temperature gradients in the atmosphere refract sound: a downward-
decreasing temperature (the normal daytime profile) bends sound
upward, away from the ground; an inversion (warm air above cool air,
common at night and in valleys) bends sound downward, toward the
ground. The downward-bending case can produce a sound shadow far
from the source and a strong audible signal at a particular distance,
the geometry of which is the standard ray-tracing problem in
atmospheric acoustics.

Wind gradients refract sound similarly: sound travels further
downwind (where the wind speed adds to the speed of sound) than
upwind. The combination of temperature and wind effects determines
the audible reach of any outdoor source.

The two refractions combine into the everyday observation that distant
sound is louder at night and downwind. By day the ground is warmer than the
air above it, the sound-speed gradient bends rays upward, and a listener far
from the source sits in an acoustic shadow that geometric spreading and
attenuation alone would not produce. At night the ground cools, the
temperature inversion bends rays back down toward the surface, and the same
source that was inaudible by day carries for kilometres. A downwind
receiver gains a second downward bending from the wind gradient, while an
upwind receiver loses the source into a shadow even at a modest range. The
practical lesson for an outdoor noise assessment is that a single
measurement is a single weather state: a community-noise prediction made
for a calm afternoon understates the night-time downwind level that
actually wakes a neighbour, and the standard environmental models carry a
meteorological correction precisely because the propagation, not the
source, has changed.

\subsection{Underwater propagation}

In water, attenuation is dominated by molecular absorption due to
magnesium sulphate and boric acid, both at $\sim\text{kHz}$
frequencies. At $1\,\si{\kilo\hertz}$ in seawater, the attenuation is
$\sim 0.07\,\si{\decibel\per\kilo\meter}$; at $10\,\si{\kilo\hertz}$,
$\sim 1\,\si{\decibel\per\kilo\meter}$; at $100\,\si{\kilo\hertz}$,
$\sim 30\,\si{\decibel\per\kilo\meter}$. Low frequencies propagate
extraordinary distances in the ocean; the SOFAR channel, a depth
band where the sound speed has a minimum due to the competition
between temperature and pressure effects, traps low-frequency sound
and lets it propagate thousands of kilometres.

The SOFAR channel is the canonical case of acoustic waveguiding by a
refracting medium. Submarines, military sonar systems, and whales
all use it. Whales' low-frequency songs at $\sim 20\,\si{\hertz}$
can be detected at intercontinental ranges in good conditions.

The waveguide forms because the sound speed in the ocean has a depth
minimum. Near the surface, warm water raises the speed; with increasing
depth the temperature falls and the speed drops, until at a few hundred
to a thousand metres the temperature stabilises and the rising pressure
takes over, lifting the speed again. A ray launched near the
speed-minimum depth bends back toward that depth from both directions by
Snell's law, because it always refracts toward the lower-speed region, so
it oscillates about the channel axis without ever reaching the lossy
surface or seabed. Energy trapped in the channel spreads cylindrically
($10\log_{10} r$) rather than spherically ($20\log_{10} r$) once it is
confined, which is the geometric reason a $20\,\si{\hertz}$ source loses
ten decibels less every decade of range than a free-field source and so
reaches across an ocean basin. The same waveguide carried the signals of
the Cold War SOSUS hydrophone arrays and today carries the long-range
calls that let a population of whales stay acoustically connected across
a basin.

\subsection{Propagation in solids and the ground}

Seismic waves in the ground are a combination of P, S, and Rayleigh
waves, each with its own velocity and attenuation. The P-wave is
the fastest and arrives first; the S-wave follows at $\sim 60\,\%$
of the P-wave speed; the Rayleigh wave at $\sim 90\,\%$ of the
S-wave speed travels along the surface and carries most of the
ground motion that buildings experience in an earthquake. The
seismometer's first measurement is the P-wave arrival; the most
damaging arrivals are the S- and Rayleigh waves that arrive seconds
later. Chapter 15 of Volume X (climate, ecological, planetary
failure) takes up earthquake engineering in depth.

The arrival-time gap between the P-wave and the S-wave is the basis of
earthquake early warning and of single-station distance estimation. Since
the P-wave runs at roughly $6\,\si{\kilo\meter\per\second}$ in crustal
rock and the S-wave at roughly $3.5\,\si{\kilo\meter\per\second}$, the
separation between the two arrivals grows with distance from the source at
a rate of about one second of P-to-S delay per eight kilometres of epicentral
distance. A station that detects the fast, low-amplitude P-wave can
estimate the distance to the rupture and broadcast a warning that reaches
distant cities before the slow, destructive S- and surface-wave train, the
seconds-to-tens-of-seconds lead that automated systems use to stop trains
and close valves. The same waves carry engineering information at smaller
scale. In near-surface geophysics, the speed of the Rayleigh wave varies
with the frequency because different frequencies sample different depths (a
long-wavelength surface wave penetrates deeper than a short one), and
inverting this frequency dependence (the dispersion curve) recovers the
shear-wave velocity profile of the ground, a standard site-investigation
method for foundation design. Ground-borne vibration from rail lines,
construction, and heavy machinery propagates by the same wave family, and
its control (trenches, wave barriers, base isolation) is the
structural-acoustics problem of section~7.7 transposed from air to soil:
the same source-path-receiver accounting applies, with the soil's
frequency-dependent attenuation replacing the air's.

\section{Resonance, room acoustics, modal density}\pathtag{standard}

A bounded volume supports standing waves at a discrete set of
natural frequencies, the room modes. The one-dimensional prototype is
a rigid tube of length $L$ closed at both ends, and it is worth solving
in full because the three-dimensional formula is the same argument run on
three axes at once.

Seek a separable solution of the one-dimensional wave equation of the
form $p(x,t) = X(x)\,T(t)$. Substituting into
$\partial_{tt}p = c^2\partial_{xx}p$ and dividing by $XT$ gives
\[
\frac{\ddot T}{c^2 T} = \frac{X''}{X},
\]
a function of $t$ alone equal to a function of $x$ alone, so both sides
equal a constant, which the oscillatory physics fixes as negative,
$-k^2$. The space factor then obeys $X'' + k^2 X = 0$, the Helmholtz
equation in one dimension, with general solution $X = a\cos(kx) +
b\sin(kx)$. The rigid walls set the boundary conditions. A rigid wall
forces zero normal particle velocity, and through the momentum relation
$\rho_0\partial_t u = -\partial_x p$ that is zero pressure gradient,
$X'(0) = X'(L) = 0$, a Neumann condition. The condition at $x=0$ kills
the sine term, $b=0$; the condition at $x=L$ requires $\sin(kL) = 0$,
hence $kL = n\pi$ for integer $n$. The allowed wavenumbers are quantised,
$k_n = n\pi/L$, the mode shapes are $p_n(x) = \cos(n\pi x/L)$ (pressure
antinodes at the rigid walls, as a hard boundary demands), and with
$k = \omega/c = 2\pi f/c$ the natural frequencies form the harmonic
ladder
\[
f_n = \frac{n c}{2L},
\]
shown in figure~\ref{fig:vol03ch07:standing-wave}. The half-wavelength of
the lowest mode fits exactly across the tube; each higher mode adds a
half-wavelength. The same ladder governs an organ pipe, an exhaust
resonator, and the lowest axial mode of any room. An open end reverses
the boundary condition to a pressure node ($X = 0$), which is why an
open-closed pipe sounds only the odd harmonics at $f_n = (2n-1)c/(4L)$
and an octave lower than the same pipe closed at both ends, the
distinction between a stopped and an open organ rank.

\input{volumes/03-force/ch07-acoustics/figures/fig-standing-wave.tex}

The three-dimensional case is the same separation run on each axis. In a
rectangular room of
dimensions $L_x \times L_y \times L_z$ with rigid walls, the modes
are
\[
f_{n_x, n_y, n_z} = \frac{c}{2}\sqrt{\left(\frac{n_x}{L_x}\right)^2
+ \left(\frac{n_y}{L_y}\right)^2 + \left(\frac{n_z}{L_z}\right)^2},
\]
with non-negative integers $n_x, n_y, n_z$ (not all zero). The first
few modes are axial (one non-zero index, the others zero), then
tangential (two non-zero), then oblique (all three non-zero). Below
the \engterm{Schroeder frequency} (typically $\sim 100$ to $300\,\si{
\hertz}$ in a domestic room), the modes are sparse and individual
modes dominate the response; above the Schroeder frequency, the
modes are densely packed and statistical methods (geometric
acoustics, ray tracing, energy decay) replace modal analysis.

\input{volumes/03-force/ch07-acoustics/figures/fig-room-modes.tex}

The transition is visible in the cumulative mode count of
figure~\ref{fig:vol03ch07:room-modes}. At low frequency the count is a
jagged staircase: each individual axial or tangential mode is audible
as a peak in the room's response, and a listener moving through the
room hears bass notes swell and null as they cross pressure antinodes
and nodes. The mode count climbs as $f^3$ (the leading Weyl term
$N(f)\approx(4\pi/3)(V/c^3)f^3$), and once the modes overlap within
their bandwidths the response smooths into the statistical regime. For
the $5\times4\times2.5\,\si{\meter}$ room of
\texttt{code/room\_modes.py}, sixty-five modes lie below
$200\,\si{\hertz}$: eleven axial, thirty-two tangential, twenty-two
oblique. The lowest is the $(1,0,0)$ axial mode at $34.3\,\si{\hertz}$,
the half-wave fit across the longest dimension.

\begin{example}[The lowest modes of a small room]
Take the same room, $L_x=5\,\si{\meter}$, $L_y=4\,\si{\meter}$,
$L_z=2.5\,\si{\meter}$, at $c=343\,\si{\meter\per\second}$. The three
lowest axial modes follow $f = c/(2L)$ along each axis:
$f_{(1,0,0)} = 343/10 = 34.3\,\si{\hertz}$,
$f_{(0,1,0)} = 343/8 = 42.9\,\si{\hertz}$,
$f_{(0,0,1)} = 343/5 = 68.6\,\si{\hertz}$. The lowest tangential mode
mixes the two longest axes:
\[
f_{(1,1,0)} = \frac{343}{2}\sqrt{\left(\tfrac{1}{5}\right)^2 +
\left(\tfrac{1}{4}\right)^2} = \frac{343}{2}\sqrt{0.04 + 0.0625}
= 54.9\,\si{\hertz}.
\]
The design lesson is in the ratios. When two room dimensions share a
common factor, their axial modes pile onto the same frequency: a cubic
room puts all three fundamental axials at one frequency, producing a
single overpowering boom and a comb of silent gaps. Spreading the
dimensions in a ratio with no small integer commensurability (the
common $1.6 : 1.25 : 1$ Bolt-area target) distributes the low modes as
evenly as the geometry allows, which is the cheapest acoustic
treatment a room can have.
\end{example}

The \engterm{reverberation time} $T_{60}$ is the time required for
the sound pressure level in a room to decay by $60\,\si{\decibel}$
after the source stops. The classic Sabine formula gives
\[
T_{60} = 0.161\,\frac{V}{A},
\]
with $V$ the room volume in $\si{\meter\cubed}$ and $A$ the total
sound absorption in $\si{\meter\squared}$-sabin, where the
sound absorption of a surface is its area times its absorption
coefficient $\alpha$ (a number between $0$, perfect reflection, and
$1$, perfect absorption). The formula is accurate in rooms with
roughly uniform absorption distribution; in rooms with patchy
absorption, the Eyring formula or numerical ray tracing is preferred.

Typical $T_{60}$ values: a concert hall at $500\,\si{\hertz}$, $\sim
2\,\si{\second}$; a recording studio, $\sim 0.3\,\si{\second}$; a
typical living room, $\sim 0.5\,\si{\second}$; an anechoic chamber,
$\sim 0.05\,\si{\second}$; a stone cathedral, $\sim 8\,\si{\second}$
or more. The optimal $T_{60}$ depends on the use: short for speech
intelligibility (lecture halls, conference rooms), longer for music
(orchestral halls, churches), with the trade-off settled by the
hall's design programme.

In practice $T_{60}$ is not read from a single $-60\,\si{\decibel}$
crossing of a noisy decay but from a straight line fitted to the
backward-integrated energy decay over a clean range and extrapolated,
as figure~\ref{fig:vol03ch07:rt60-decay} shows. The standard fit range
is $-5$ to $-35\,\si{\decibel}$ (the $T_{30}$ estimate, scaled by two),
chosen to avoid both the early-reflection curvature near the top and
the noise floor near the bottom. Fitting over a defined range is what
makes the reverberation time repeatable between measurements and
between analysts.

\input{volumes/03-force/ch07-acoustics/figures/fig-rt60-decay.tex}

\begin{example}[Sabine prediction for a furnished room]
For the $50\,\si{\meter\cubed}$ room, take surface absorptions at
$500\,\si{\hertz}$: a carpeted floor of $20\,\si{\meter\squared}$ at
$\alpha=0.30$, a plaster ceiling of $20\,\si{\meter\squared}$ at
$\alpha=0.05$, painted walls of $45\,\si{\meter\squared}$ at
$\alpha=0.06$, furniture and drapery contributing
$8\,\si{\meter\squared}$-sabin equivalent at $\alpha=0.50$, and
$3\,\si{\meter\squared}$ of glass at $\alpha=0.18$. The total
absorption is
\[
A = 20(0.30) + 20(0.05) + 45(0.06) + 8(0.50) + 3(0.18)
  \approx 14.2\,\si{\meter\squared}\text{-sabin},
\]
and the Sabine reverberation time is
\[
T_{60} = 0.161\,\frac{V}{A} = 0.161\,\frac{50}{14.2}
       \approx 0.57\,\si{\second},
\]
a value squarely in the living-room range. The carpet and the soft
furnishings carry most of the absorption; the hard ceiling and walls
carry almost none, which is why adding a rug or curtains audibly
shortens a room's reverberation while painting the walls does nothing.
\texttt{code/sabine\_rt60.py} reproduces this calculation and the
Eyring correction, which lowers the estimate to
$\approx 0.52\,\si{\second}$ once the mean absorption is no longer
small.
\end{example}

\begin{estimation}
\textbf{Question.} A bare concrete parking garage, roughly
$30\,\si{\meter}\times20\,\si{\meter}\times3\,\si{\meter}$, sounds
cavernous. Estimate its reverberation time, then estimate how much
absorptive area would be needed to bring it to a speech-friendly
$T_{60}\approx 1\,\si{\second}$.

\textbf{Setup.} The volume is $V \approx 30\times20\times3 =
1800\,\si{\meter\cubed}$. The total surface is two floors-and-ceiling
of $600\,\si{\meter\squared}$ each plus four walls totalling roughly
$2(30+20)\times3 = 300\,\si{\meter\squared}$, so
$S \approx 1500\,\si{\meter\squared}$. Bare sealed concrete has an
absorption coefficient of order $\alpha \approx 0.02$ across the speech
band.

\textbf{Bare garage.} The absorption is
$A \approx 1500\times 0.02 = 30\,\si{\meter\squared}$-sabin, giving
\[
T_{60} \approx 0.161\,\frac{1800}{30} \approx 9.7\,\si{\second}.
\]
Nearly ten seconds of reverberation, which matches the lived
experience of a concrete garage: speech is unintelligible because each
syllable is still ringing when the next arrives.

\textbf{Target.} For $T_{60}\approx 1\,\si{\second}$ the required
absorption is $A = 0.161\,V/T_{60} \approx 0.161\times 1800/1 \approx
290\,\si{\meter\squared}$-sabin. Subtracting the
$30\,\si{\meter\squared}$-sabin the concrete already provides leaves
$\sim 260\,\si{\meter\squared}$-sabin to add. With a good broadband
absorber at $\alpha\approx 0.8$, that is $\sim 325\,\si{\meter\squared}$
of treated area, roughly a fifth of the total surface, plausibly the
whole ceiling. The estimate sets the order of magnitude of the
intervention: treating one large surface, not scattering a few panels,
is what a tenfold reduction in reverberation costs.
\end{estimation}

\section{Ultrasonic and infrasonic engineering}\pathtag{standard}

\engterm{Ultrasonic} frequencies, above the human audible range
($\gtrsim 20\,\si{\kilo\hertz}$), are an engineering tool. The
short wavelength makes the propagation behave nearly geometrically,
which enables imaging, ranging, and inspection. Standard
engineering uses:

\begin{itemize}[leftmargin=*]
  \item \engterm{Medical ultrasound imaging} ($1$ to $15\,\si{\mega
    \hertz}$). Pulses of ultrasound enter the body; echoes from
    tissue interfaces (skin-fat, fat-muscle, muscle-bone) are
    measured and time-of-flight gives depth. Two-dimensional B-mode
    images come from raster-scanning the beam direction.
  \item \engterm{Ultrasonic non-destructive testing} ($0.5$ to
    $25\,\si{\mega\hertz}$). Echoes from internal defects (cracks,
    voids, inclusions) in metals reveal flaws invisible from
    outside. Standard in pressure-vessel inspection, weld testing,
    and rotating-component fatigue inspection (chapter 12 of Volume
    X).
  \item \engterm{Sonar} ($1$ to $500\,\si{\kilo\hertz}$). Active
    pulse-echo systems on ships and submarines map underwater
    topology, detect other vessels, and measure water depth. Passive
    sonar listens for noise radiated by other vessels.
  \item \engterm{Ultrasonic cleaning} ($\sim 40\,\si{\kilo\hertz}$).
    Cavitation in liquid baths driven by powerful ultrasonic
    transducers dislodges particulate contamination from immersed
    parts.
  \item \engterm{Ultrasonic welding} ($\sim 20$ to $40\,\si{\kilo
    \hertz}$). High-amplitude vibration of a horn against two
    pieces of thermoplastic generates frictional heat that fuses
    them.
\end{itemize}

\engterm{Infrasonic} frequencies, below the human audible range
($\lesssim 20\,\si{\hertz}$), are also engineering tools. Long
wavelengths penetrate obstacles and travel long distances. Standard
uses:

\begin{itemize}[leftmargin=*]
  \item \engterm{Infrasound monitoring} of nuclear-test compliance
    and large-scale natural phenomena (volcanic eruptions, large
    explosions, severe weather). The Comprehensive Nuclear-Test-Ban
    Treaty's International Monitoring System includes a global
    network of $60$ infrasound stations.
  \item \engterm{Seismic monitoring} of earthquakes, with the lowest
    frequencies of structural-engineering interest in the $0.1$ to
    $10\,\si{\hertz}$ band (the band that excites buildings'
    fundamental modes).
  \item \engterm{Wind-turbine noise monitoring}: large wind turbines
    radiate low-frequency sound from blade-pass and gearbox
    interactions, in the $0.5$ to $50\,\si{\hertz}$ band; community
    annoyance and health effects are an active design constraint.
\end{itemize}

Infrasound monitoring works because the low frequencies the atmosphere
carries best are exactly the ones large events radiate. The classical and
relaxation absorption that kills a kilohertz tone in a few kilometres leaves
a one-hertz wave almost untouched, since the absorption falls as the square
of frequency, so a $1\,\si{\hertz}$ signal survives to global range with an
atmospheric loss measured in small fractions of a decibel per thousand
kilometres. A large explosion, a volcanic eruption, or a bolide couples its
energy into this band and the wave circles the globe; the $1883$ Krakatoa
eruption drove a pressure pulse that barographs recorded on several
passages around the Earth, and the same physics lets the treaty-monitoring
network of infrasound arrays detect an atmospheric nuclear test at
continental range from the arrival-time difference across an array of
microbarometers. The detection problem is not attenuation but the wind and
temperature noise that fills the same low band, which is why an infrasound
station is not a single sensor but an array of microbarometers spread over
tens to hundreds of metres, each fed through a spatial filter (a rosette of
porous pipes) that averages out the small-scale turbulent pressure while
passing the coherent long-wavelength signal, the spatial analogue of the
matched filter that pulls a coded pulse out of noise.

The wind-turbine case is where infrasound becomes a design constraint rather
than a monitoring opportunity. Each blade passes the tower once per
revolution, so a three-blade rotor turning at a fraction of a hertz radiates
a blade-pass tone and its harmonics in the low single-digit hertz, below the
threshold of clear hearing but within the range at which the pulsating
pressure is felt and at which building structures have their own low modes.
The engineering response is the source-path-receiver accounting of the next
section applied at frequencies where a barrier is useless (a $2\,\si{\hertz}$
wavelength is $170\,\si{\meter}$, so nothing short of a hill diffracts it)
and where the only effective levers are at the source: blade-tower clearance,
rotor speed, and pitch scheduling that smooth the pressure pulse each blade
delivers as it passes the tower.

Medical ultrasound imaging deserves the mechanism in full, because it makes
the chapter's impedance and attenuation ideas visible in a device millions
of people meet. An image forms because tissue interfaces reflect: at a
boundary between two tissues of impedances $Z_1$ and $Z_2$ the echo carries a
fraction $[(Z_2-Z_1)/(Z_2+Z_1)]^2$ of the incident intensity, and soft
tissues differ in impedance by only a few percent, so each interface returns
a faint echo and most of the pulse continues deeper to sound the next layer.
That weak contrast is a feature: a strong reflector would return everything
at the first interface and leave nothing to image below it, which is exactly
why a bone or a gas bubble ($Z$ mismatched by orders of magnitude) casts an
acoustic shadow, and why a coupling gel excludes the air layer that would
otherwise reflect the whole pulse at the skin. The design tension is
frequency. Axial resolution improves with frequency because a higher
frequency permits a shorter pulse, and lateral resolution improves because a
larger $ka$ focuses a tighter beam, but tissue attenuation climbs at roughly
$0.5\,\si{\decibel}$ per centimetre per megahertz, so a $10\,\si{\mega\hertz}$
probe that resolves a tenth of a millimetre penetrates only a few centimetres
while a $2\,\si{\mega\hertz}$ probe reaches deep abdominal organs at coarser
resolution. Every probe frequency is a depth-versus-resolution choice, the
same trade the general pulse-echo discussion states, sharpened here by a
specific attenuation law. The instrument compensates the near-to-far echo
decay with a time-gain amplifier that boosts later (deeper) echoes to
equalise the brightness of equal reflectors at different depths, and it
suppresses the transducer's own ring, the reason the quarter-wave matching
layer of the worked example matters, so a faint deep echo is not buried under
the emitter's afterglow.

Two parameters govern any pulse-echo ranging system, and the ultrasonic
imaging, NDT, and sonar uses above are the same physics at different
frequencies and in different media. The first is the maximum range, set by
how far a pulse of given energy travels before its echo falls below the
noise floor, which the attenuation curves of section~7.4 make a steep
function of frequency: a $100\,\si{\kilo\hertz}$ sonar that ranges a few
kilometres becomes a few-hundred-metre instrument at $500\,\si{\kilo\hertz}$
because the seawater absorption climbs. The second is the range resolution,
the smallest separation between two reflectors whose echoes can still be
told apart, which equals half the spatial pulse length $c\tau/2$ for a pulse
of duration $\tau$. The two pull against each other: a short pulse resolves
finely but carries little energy and so ranges poorly, while a long pulse
ranges far but smears nearby reflectors together. Modern systems escape the
trade with a coded long pulse (a chirp or a pseudo-random sequence)
compressed on reception by matched filtering, which recovers the resolution
of a short pulse while keeping the energy of a long one, the same
pulse-compression trick radar uses. Higher frequency buys finer resolution
through both the shorter achievable pulse and the narrower beam of the
$ka$ relation below, at the price of range, which is the single trade that
sets the operating frequency of every imaging and ranging device in the
chapter.

The engineering of acoustic transducers (microphones, hydrophones,
ultrasonic probes, accelerometers operating as acoustic sensors) is
the bridge between the wave's mechanical behaviour and the
electrical signal the engineer can record. Chapter 9 of Volume VII
takes up transducer design from the signal-and-information side.

A radiator's \engterm{directivity} sets how the energy spreads in
angle, and it is governed by the size parameter $ka = 2\pi a/\lambda$,
the radiator radius measured in wavelengths.
Figure~\ref{fig:vol03ch07:directivity} contrasts a small radiator
($ka\approx 2$), which radiates almost omnidirectionally, with a large
one ($ka\approx 8$), which concentrates its energy into a narrow
forward lobe. The exact pattern of a flat circular piston is
$|2J_1(ka\sin\theta)/(ka\sin\theta)|$, whose main-lobe half-width
scales as $\sim\lambda/(2a)$. The trade is unavoidable at fixed
frequency: a narrow beam needs a large aperture. A sonar that must
resolve a small target needs either a high frequency (short
wavelength) or a large array; a tweeter that must spread sound across
a room needs to stay small compared to the wavelengths it radiates.
The same relation sets the resolution of an ultrasound image and the
beamwidth of a public-address horn.

\input{volumes/03-force/ch07-acoustics/figures/fig-directivity.tex}

The plane-wave picture that carried every duct and muffler result holds
only while the duct is narrow compared to a wavelength. Above a geometry-set
cutoff a duct behaves as a waveguide, its higher modes propagate
dispersively, and the one-dimensional acoustics of the muffler examples
breaks down. A rigid rectangular ventilation duct is
$0.30\,\si{\meter}$ wide. Find the frequency above which its first
higher-order cross-mode propagates, and the phase and group velocities of
that mode at $1.5$ times cutoff.

\textbf{The cutoff frequency.} A rigid-walled duct of width $b$ supports a
cross-mode with a pressure half-wavelength across the duct, which propagates
only above the cutoff frequency
\[
f_c = \frac{c}{2b} = \frac{343}{2(0.30)} \approx 572\,\si{\hertz}.
\]
Below $572\,\si{\hertz}$ only the plane-wave mode carries energy; the
cross-mode is evanescent, decaying along the duct without transport, and
one-dimensional plane-wave silencer theory applies. Above it the cross-mode
propagates and the field is genuinely two-dimensional.

\textbf{Dispersion above cutoff.} A propagating higher mode has an axial
wavenumber $k_z = \sqrt{(\omega/c)^2 - (\pi/b)^2}$, which makes its phase and
group velocities frequency-dependent, the two curves of
figure~\ref{fig:vol03ch07:duct-cutoff}. At $f = 1.5 f_c$,
\[
c_{\text{ph}} = \frac{c}{\sqrt{1-(f_c/f)^2}}
             = \frac{343}{\sqrt{1-(1/1.5)^2}}
             \approx 461\,\si{\meter\per\second},
\]
faster than $c$, while the energy travels at the group velocity
\[
c_{\text{gr}} = c\sqrt{1-(f_c/f)^2}
             = 343\sqrt{1-(1/1.5)^2}
             \approx 256\,\si{\meter\per\second},
\]
slower than $c$, their product recovering $c_{\text{ph}}c_{\text{gr}}=c^2$.

\textbf{Why the cutoff matters to a silencer.} A dissipative silencer
designed on plane-wave attenuation loses accuracy above its duct cutoff,
because a higher mode can carry energy down the centre of the passage where
the wall lining reaches it weakly, a leakage path the one-dimensional
calculation cannot see. Splitting a wide duct with a central baffle, the
trick the ventilation-silencer design exercise uses, raises the effective
cutoff by halving the passage width, which is as much about suppressing the
higher modes as about doubling the lined perimeter. The same cutoff sets the
highest frequency at which a muffler's four-pole model of the preceding
mastery box stays valid.

\input{volumes/03-force/ch07-acoustics/figures/fig-duct-cutoff.tex}

\begin{mastery}
\textbf{Numerical acoustics: FEM and BEM.} Closed-form room modes and
piston directivities exist only for the simplest geometries. Real
cabins, mufflers, concert halls, and transducer housings are solved
numerically, and the choice between the two dominant methods turns on
where the unknowns live. The \engterm{finite element method} (FEM)
discretises the volume of the acoustic domain into elements and solves
the Helmholtz equation $\nabla^2 p + k^2 p = 0$ over the mesh; it suits
bounded interior problems (a car cabin, a muffler, a room) and couples
naturally to a structural FEM model for vibro-acoustic problems where
the walls flex. Its cost grows with the volume meshed, and it needs an
artificial absorbing boundary (a perfectly matched layer) to fake an
open domain. The \engterm{boundary element method} (BEM) discretises
only the surfaces and uses the free-space Green's function to satisfy
the radiation condition at infinity automatically; it suits exterior
radiation and scattering problems (sound radiated from a vibrating
machine, scattering off a barrier) where meshing the surrounding air
would be wasteful. Its matrices are dense rather than sparse, so it
scales worse with surface size unless accelerated (the fast multipole
method). Both methods share the engineering hazard of all wave
computation: the mesh must resolve the wavelength, conventionally six
to ten elements per wavelength, so the element count climbs as
$f^3$ for a volume method and the high-frequency end of any band is the
expensive end. Above the Schroeder frequency the deterministic methods
give way to statistical energy analysis, which tracks band-averaged
energy flow between subsystems rather than the field itself.
\end{mastery}

\section{Noise control: sources, paths, receivers}\pathtag{core}

Noise is unwanted sound. Engineering noise control follows the
\engterm{source-path-receiver} model:

\engterm{Source}. Reduce the noise at the source. This is the most
effective intervention but also the most expensive: a quieter
machine costs more than a louder one. Sources include rotating
machinery (chapter 4 of this volume), fluid flow (turbulence in
ducts, valve cavitation, fan blade-pass tones), combustion (engine
exhaust, jet engines), impact (presses, hammers, gears), and
aerodynamic flow over structures (wind on vehicles, propellers).

\engterm{Path}. Reduce the transmission from source to receiver. The
canonical interventions are barriers (walls), enclosures (boxes),
acoustic absorbers (foam, fibreglass, perforated panels backed by
absorber), and mufflers/silencers (reactive and dissipative
elements in flow paths). Each is tuned to the dominant frequency
band of the noise: low-frequency noise requires mass (heavy
barriers) and low-frequency absorbers (deep porous absorbers,
Helmholtz resonators); high-frequency noise responds to lighter
barriers and shallow absorbers.

\engterm{Receiver}. Reduce the noise reaching the receiver, typically
by hearing protection (earplugs, earmuffs), active noise-cancellation
headsets, or relocating the receiver. Hearing protection is the
last line of defence; it is cheap but depends on consistent use
and provides $\sim 15$ to $30\,\si{\decibel}$ of attenuation at
audio frequencies.

\engterm{Noise standards}. OSHA and EU directive 2003/10/EC set
occupational noise exposure limits in the workplace (typically
$85\,\si{\decibel}$ A-weighted for an $8$-hour day, with reduced
limits for higher exposure levels). Environmental noise standards
(road traffic, aircraft, industrial sources) are set by national
and local regulators; the World Health Organization recommends
$45\,\si{\decibel}$ A-weighted as the night-time outdoor limit for
sleep quality. Each standard requires the engineer to measure the
noise at a specified location and to compare to a band-weighted
limit.

\engterm{A-weighting}. The human auditory system is more sensitive
to mid-frequency sound (around $1$ to $4\,\si{\kilo\hertz}$) than to
low- or high-frequency sound at the same SPL. The A-weighting filter
applies a frequency-dependent gain that approximately matches human
sensitivity at moderate sound levels; the A-weighted level
$\text{SPL}_A$ in dB(A) is the standard measure for occupational and
environmental noise. Other weightings (B, C, D, Z) exist for
specific applications (impulsive noise, aircraft, unweighted
measurement).

\input{volumes/03-force/ch07-acoustics/figures/fig-weighting-curves.tex}

The weighting is a fixed frequency response applied before the spectrum
is summed. Figure~\ref{fig:vol03ch07:weighting-curves} plots the A and C
curves, both standardised so the gain is zero at $1\,\si{\kilo\hertz}$. To
A-weight a measured octave-band spectrum the engineer adds the band's A
offset to its level and energy-sums the result by the rule above; the C
curve, nearly flat across the audible band and rolling off only at the
extremes, is the right tool when the low-frequency energy is a real load,
as for a peak level or an impulsive event. The A and C levels of the same
noise differ by an amount that itself diagnoses the spectrum: a large
C-minus-A difference signals a noise dominated by bass content that
A-weighting has discarded, the warning sign that a structural or a
low-frequency-annoyance problem hides behind a comfortable A-weighted
number.

\input{volumes/03-force/ch07-acoustics/figures/fig-spl-octave.tex}

Figure~\ref{fig:vol03ch07:spl-octave} shows a machine's octave-band
spectrum before and after A-weighting. The unweighted spectrum is
dominated by a low-frequency band that A-weighting discounts heavily,
so the single-number A-weighted level sits well below the unweighted
energy sum and is shifted toward the mid bands the ear hears best.
\texttt{code/spl\_octave.py} reads the band levels from
\texttt{data/octave\_spectrum.csv}, applies the standard octave-band
A-weighting offsets, and returns $90.2\,\si{\decibel}$ unweighted and
$81.4\,\si{\decibel}$ A-weighted: an $8.8\,\si{\decibel}$ reduction
that lands the machine just below the typical $85\,\si{\decibel}$(A)
occupational limit. The same calculation shows where treatment pays:
the $63\,\si{\hertz}$ band that dominates the unweighted level barely
registers after weighting, so spending money to quiet it would satisfy
a physicist and not an inspector. The band that controls the
A-weighted total is the one to treat.

\subsection{Case study: a noise-control enclosure carried end to end}

The source-path-receiver model becomes an engineering calculation only
when the levels are carried through it with numbers. We work one
enclosure from the machine's radiated power to the worker's daily dose
and a pass-or-fail verdict against an exposure limit, so that every
stage of the model produces a quantity the next stage consumes.

\textbf{The problem.} A stamping press sits on the hard concrete floor
of a fabrication shop. A worker operates it from a station
$2\,\si{\meter}$ away for a full shift. The plant safety officer
measures $98\,\si{\decibel}$(A) at the station and asks whether an
enclosure can bring the worker's eight-hour exposure below the
$85\,\si{\decibel}$(A) action level of the occupational standard, and
if so, what the enclosure must be.

\textbf{Step 1: source sound power.} The first move is to convert the
station level into a source-intrinsic quantity, because the enclosure
will change the field but not the machine. Treating the press as
radiating into the half-space above the floor, a measurement hemisphere
of radius $r=2\,\si{\meter}$ has area $S=2\pi r^2 = 25.1\,\si{
\meter\squared}$. The surface-averaged level near the station is taken
as $\overline{L_p}\approx 98\,\si{\decibel}$(A), so the radiated sound
power level is
\[
L_W = \overline{L_p} + 10\log_{10}\frac{S}{S_0}
    = 98 + 10\log_{10}(25.1) \approx 98 + 14 = 112\,\si{\decibel}\text{(A)},
\]
referenced to $S_0=1\,\si{\meter\squared}$ and a $1\,\si{\pico\watt}$
power reference. The press radiates about $0.16\,\si{\watt}$ of acoustic
power, a number that stays fixed as we build the box around it.

\textbf{Step 2: required transmission loss.} The dose target sets the
allowed receiver level. Under the equal-energy ($3\,\si{\decibel}$)
exchange rule, a full eight-hour exposure must not exceed
$85\,\si{\decibel}$(A); we design to $82\,\si{\decibel}$(A) to leave a
$3\,\si{\decibel}$ margin against measurement and fit uncertainty. The
untreated station level is $98\,\si{\decibel}$(A), so the enclosure must
deliver an insertion loss of at least
\[
\text{IL}_{\text{req}} = 98 - 82 = 16\,\si{\decibel}\text{(A)}.
\]
The dominant content of a stamping press is mid-band impact noise; we
size the panel for the $500\,\si{\hertz}$ octave and require the
transmission loss there to clear the insertion-loss target with room for
the losses that follow.

\textbf{Step 3: panel mass law.} A single limp panel blocks sound by its
inertia. The field-incidence mass law is
\[
\text{TL} \approx 20\log_{10}(m'' f) - 47\,\si{\decibel},
\]
with $m''$ the surface mass in $\si{\kilogram\per\meter\squared}$ and
$f$ in $\si{\hertz}$, plotted across frequency in
figure~\ref{fig:vol03ch07:mass-law-tl}. Steel of $1.8\,\si{\milli\meter}$
thickness has $m''=14\,\si{\kilogram\per\meter\squared}$, giving at
$500\,\si{\hertz}$
\[
\text{TL} = 20\log_{10}(14\times 500) - 47
          = 20\log_{10}(7000) - 47 \approx 76.9 - 47 \approx 30\,\si{\decibel},
\]
the design point marked on the figure. The bare panel offers nearly
$30\,\si{\decibel}$, comfortably above the $16\,\si{\decibel}$ target,
which tempts the inexperienced designer to declare victory. The
remaining steps show why the panel TL is not the insertion loss.

\textbf{Step 4: leakage and flanking.} An enclosure is its weakest path.
The composite transmission loss of a wall that is mostly a good panel of
area $S_w$ with transmission coefficient $\tau_w$ but has a small gap or
weak element of area $S_g$ with $\tau_g$ is set by the area-weighted
transmission coefficient
\[
\bar\tau = \frac{S_w\tau_w + S_g\tau_g}{S_w + S_g},
\qquad \text{TL}_{\text{comp}} = 10\log_{10}\frac{1}{\bar\tau}.
\]
Take a $30\,\si{\decibel}$ panel ($\tau_w=10^{-3}$) of
$S_w=20\,\si{\meter\squared}$ with an unsealed door gap of
$S_g=0.02\,\si{\meter\squared}$ that transmits freely ($\tau_g\approx 1$).
Then
\[
\bar\tau = \frac{20(10^{-3}) + 0.02(1)}{20.02}
         = \frac{0.020 + 0.020}{20.02} \approx 2.0\times 10^{-3},
\]
so $\text{TL}_{\text{comp}}\approx 10\log_{10}(500)\approx 27\,\si{
\decibel}$, already three decibels below the panel alone, and a gap ten
times larger would erase half the panel's benefit. Allowing for a sealed
but imperfect door, structure-borne flanking through the floor mounts,
and the conduit and ventilation penetrations every real enclosure
carries, the achievable insertion loss is taken as
$22\,\si{\decibel}$ rather than the panel's $30$. The receiver level
after the path is
\[
98 - 22 = 76\,\si{\decibel}\text{(A)},
\]
the third bar of figure~\ref{fig:vol03ch07:enclosure-budget}.

\textbf{Step 5: absorption inside.} A bare-walled enclosure traps the
machine's own sound, the interior reverberates, the internal level
climbs, and the extra interior pressure pushes more energy through the
walls, eroding the insertion loss by several decibels. Lining the
interior with $50\,\si{\milli\meter}$ of porous absorber raises the mean
interior absorption from a bare-steel $\bar\alpha\approx 0.05$ to
$\bar\alpha\approx 0.7$. The reverberant-level reduction is
$10\log_{10}(\bar\alpha_2/\bar\alpha_1) = 10\log_{10}(0.7/0.05)\approx
11\,\si{\decibel}$ of interior buildup removed, of which a conservative
$3\,\si{\decibel}$ shows up as recovered insertion loss at the receiver.
The received level becomes
\[
76 - 3 = 73\,\si{\decibel}\text{(A)},
\]
the last bar of figure~\ref{fig:vol03ch07:enclosure-budget}.

\textbf{Step 6: daily noise dose and verdict.} The worker is exposed to
$73\,\si{\decibel}$(A) for the eight-hour shift. The dose under the
equal-energy rule, with the $85\,\si{\decibel}$(A) eight-hour criterion,
is
\[
D = \frac{T_{\text{exposure}}}{T_{\text{allowed}}}\times 100\,\%,
\qquad
T_{\text{allowed}} = \frac{8}{2^{(L-85)/3}}\,\text{h}.
\]
At $L=73\,\si{\decibel}$(A), $T_{\text{allowed}} = 8\times 2^{12/3} =
8\times 16 = 128\,\si{\hour}$, so an eight-hour exposure is
\[
D = \frac{8}{128}\times 100\,\% \approx 6\,\%,
\]
far below the $100\,\%$ that the limit permits. The enclosure passes
with wide margin: the worker could spend the entire shift at the
station, and could absorb several short excursions to the untreated
field, without reaching the dose limit. Had we stopped at the panel mass
law and claimed the full $30\,\si{\decibel}$, we would have predicted
$68\,\si{\decibel}$(A) and a $2.0\,\%$ dose; the five-decibel gap
between that optimistic figure and the realised $73\,\si{\decibel}$(A)
is the leakage-and-flanking reality that
figure~\ref{fig:vol03ch07:enclosure-budget} accounts for, and it is the
gap that turns enclosures that pass on paper into enclosures that fail
on the floor.

\input{volumes/03-force/ch07-acoustics/figures/fig-mass-law-tl.tex}

\input{volumes/03-force/ch07-acoustics/figures/fig-enclosure-budget.tex}

\textbf{What the case study fixes in habit.} The transport accounting
of the chapter's opening archetype runs straight down the budget: power
radiated at the source, attenuated along the path by mass and degraded
by leakage, partly recovered by interior absorption, and delivered to a
receiver whose tolerance is set by an energy dose rather than a peak
level. The design verdict turns on the worst path, not the best panel,
and on the dose integral, not the instantaneous level. An enclosure
specified from the panel mass law alone is a specification of the part
of the problem that was never in doubt.

\subsection{Case study: an auditorium reverberation design carried end to
end}

The enclosure case study followed energy out of a room. A hall design
follows energy that must stay in the room long enough to sound alive and
leave fast enough to keep speech intelligible. We carry one
multi-purpose auditorium from its seat count and use programme to a
surface-by-surface absorption budget, a reverberation prediction at the
mid band, an early-reflection clarity check, and a verdict against the use
target, so that the Sabine accounting of section~7.5 becomes a build
specification rather than a single formula evaluation.

\textbf{The problem.} A municipal arts centre commissions a hall for
$500$ seats serving amplified speech (lectures, conferences) and
small-ensemble and amplified music, with no requirement for unamplified
symphonic programme. The brief asks for a hall whose reverberation
supports clear speech yet does not sound dead, with no slap echo from the
rear wall and adequate early-reflection support at the rear seats. The
acoustician must size the volume, set a reverberation target, and write
the absorption budget that the finishes contractor will build to.

\textbf{Step 1: volume per seat and the geometry.} The first move is to
size the room from the seat count, because the volume sets both the
reverberation that is achievable and the air supply the seats need. A
speech-and-light-music hall is designed at roughly $7$ to
$9\,\si{\meter\cubed}$ per seat; a symphonic hall runs higher, $10$ to
$12$, to win reverberation, while a cinema runs lower. Taking
$8.4\,\si{\meter\cubed}$ per seat for $500$ seats gives a target volume
\[
V = 500\times 8.4 = 4200\,\si{\meter\cubed},
\]
realised as a fan-shaped room about $30\,\si{\meter}$ deep,
$24\,\si{\meter}$ wide at the rear, and $7\,\si{\meter}$ to the ceiling
underside, the design point of figure~\ref{fig:vol03ch07:rt60-targets}.
The total interior surface works out to roughly
$S\approx 1900\,\si{\meter\squared}$ across floor, ceiling, side walls,
rear wall, and stage opening.

\textbf{Step 2: the reverberation target.} The use programme sets the
mid-band target. Figure~\ref{fig:vol03ch07:rt60-targets} places a
$4200\,\si{\meter\cubed}$ speech-and-light-music hall at about
$1.0\,\si{\second}$ at $500\,\si{\hertz}$, a value short enough to keep
consonants distinct (a hall above $\sim 1.4\,\si{\second}$ smears speech)
and long enough to avoid the dead, fatiguing quality of an
over-damped room. We also set a bass target slightly above the mid band,
$T_{60}(125\,\si{\hertz})\approx 1.1\,\si{\second}$, the modest
warmth a music programme wants, and a treble target slightly below,
$T_{60}(2\,\si{\kilo\hertz})\approx 0.9\,\si{\second}$, accepting the air
absorption that naturally rolls off the high end in a room this size.

\textbf{Step 3: the absorption the target demands.} Inverting the Sabine
formula for the occupied hall gives the total absorption the finishes plus
the audience must supply:
\[
A_{\text{req}} = 0.161\,\frac{V}{T_{60}}
              = 0.161\,\frac{4200}{1.0}
              \approx 676\,\si{\meter\squared}\text{-sabin}
\]
at $500\,\si{\hertz}$. This is the budget the room must hit when full.

\textbf{Step 4: the audience is the largest absorber.} A seated audience
dominates the budget, which is the controlling fact of large-hall design.
An occupied upholstered seat absorbs about
$\alpha\approx 0.85$ per seat-area at $500\,\si{\hertz}$, and the
audience-plus-seating block of $500$ seats covers about
$0.9\,\si{\meter\squared}$ of floor each. Taking the occupied audience
area as $A_{\text{aud}}\approx 450\,\si{\meter\squared}$ at an effective
$\alpha\approx 0.85$ contributes
\[
382.5\,\si{\meter\squared}\text{-sabin},
\]
the dominant bar of figure~\ref{fig:vol03ch07:absorption-budget}. The
occupied audience alone supplies more than half the $676$ target from a
single surface, and with the finishes added the full hall runs past it,
which warns of
the central hazard: the hall will be over-damped when full and far too
live when empty unless the seats are specified to absorb almost as much
empty as occupied. Heavily upholstered seats (perforated underside,
porous fill) close most of the occupied-to-empty gap, holding the
reverberation roughly constant with occupancy. We specify such seats and
treat the remaining surfaces lightly.

\textbf{Step 5: the surface budget and the rear wall.} With the audience
carrying the bulk, the remaining surfaces are budgeted for two distinct
jobs: a small absorption top-up and echo control. The reflective ceiling
and the side walls near the stage are kept hard ($\alpha\approx 0.04$ to
$0.06$) so that they return early reflections to the seats. The rear wall,
$24\,\si{\meter}$ from the stage, is the slap-echo risk: a sound leaving
the stage, reflecting off a hard rear wall, and returning to the
performer travels about $48\,\si{\meter}$ extra, a delay of
$48/343\approx 140\,\si{\milli\second}$, well above the
$\sim 50\,\si{\milli\second}$ threshold at which a reflection is heard as
a discrete echo rather than fused with the direct sound. We treat the
rear wall with a deep porous absorber ($\alpha\approx 0.8$ over
$\sim 150\,\si{\meter\squared}$) to kill the return, contributing the
$120\,\si{\meter\squared}$-sabin rear-wall bar of
figure~\ref{fig:vol03ch07:absorption-budget}. The seats, ceiling, side
walls, floor, and stage opening add the smaller bars, and the budget
sums to
\[
A \approx 382.5 + 90 + 40 + 60 + 18 + 120 + 55
      \approx 766\,\si{\meter\squared}\text{-sabin}.
\]

\textbf{Step 6: reverberation check.} The budget modestly overshoots the
target, which means the as-built hall reverberates a little shorter than
$1.0\,\si{\second}$:
\[
T_{60} = 0.161\,\frac{4200}{766} \approx 0.88\,\si{\second}.
\]
A $0.88\,\si{\second}$ hall is fine for amplified speech and serviceable
for amplified music, but it sits just short of the $1.0\,\si{\second}$
liveness target, confirming the step-4 warning that a full audience
over-damps the room. The design correction is to remove discretionary
absorption: drop the side-wall and ceiling treatment to bare hard finish
(already assumed), shrink the rear-wall absorber to the minimum area that
still kills the echo (a row of absorptive panels at the echo-return height
rather than a full wall), and let the seats carry the decay. Pulling the
non-audience absorption down from $383$ to about
$293\,\si{\meter\squared}$-sabin, so that $A\approx 382.5+293\approx
676\,\si{\meter\squared}$-sabin, recovers
\[
T_{60} = 0.161\,\frac{4200}{676} \approx 1.0\,\si{\second},
\]
landing on target. The lesson is that in a full hall the designer spends
most effort taking absorption away, not adding it, because the audience
has already supplied the budget.

\textbf{Step 7: early-reflection and clarity check.} Reverberation time
alone does not certify intelligibility. A rear-seat listener
$28\,\si{\meter}$ from the stage receives the direct sound after
$28/343\approx 82\,\si{\milli\second}$; a useful early reflection off the
ceiling or a side wall must arrive within about $50\,\si{\milli\second}$
of the direct sound to reinforce it rather than blur it. We size the
ceiling tilt so that the stage-to-ceiling-to-rear-seat path exceeds the
direct path by no more than $\sim 17\,\si{\meter}$
($50\,\si{\milli\second}\times 343$), which a ceiling sloped to about
$7\,\si{\meter}$ over the stage rising to $9\,\si{\meter}$ over the rear
seats achieves. The early-to-late energy ratio, the clarity index
$C_{80}=10\log_{10}(E_{0\text{-}80}/E_{80\text{-}\infty})$, should sit
near $0$ to $+4\,\si{\decibel}$ for a speech-and-light-music hall; the
$0.97\,\si{\second}$ reverberation with strong early ceiling reflections
puts the estimated $C_{80}$ in that band, so the hall is predicted to be
both adequately live and clear.

\textbf{Verdict.} The hall passes: $V=4200\,\si{\meter\cubed}$ at
$8.4\,\si{\meter\cubed}$ per seat, occupied mid-band
$T_{60}\approx 0.97\,\si{\second}$ against a $1.0\,\si{\second}$ target,
rear-wall echo controlled below the $50\,\si{\milli\second}$ fusion
threshold, and an early-reflection geometry that keeps the rear seats
inside the clarity window. The empty hall, with the specified
self-absorbing seats, runs only about $0.1$ to $0.2\,\si{\second}$ longer,
inside the tolerance for rehearsal use. The single transportable habit is
that hall design is an absorption budget dominated by the audience, in
which the acoustician's surfaces correct the small remainder and control
echoes, and the reverberation target is met by deciding how much
absorption to withhold rather than how much to add.

\input{volumes/03-force/ch07-acoustics/figures/fig-rt60-targets.tex}

\input{volumes/03-force/ch07-acoustics/figures/fig-absorption-budget.tex}

\subsection{Case study: an engine exhaust muffler carried end to end}

The enclosure study interrupted an airborne path and the auditorium study
shaped a reverberant field. A muffler interrupts a path that also carries a
hot, fast, pulsating flow, so the design must satisfy an acoustic target
and a back-pressure budget at once. We carry one passenger-car exhaust
muffler from the engine's firing-order spectrum through a reactive
expansion chamber, a dissipative lined section, and a flow-resistance check
against an engine-performance limit, so that the transport accounting of
the chapter runs down a path where the medium itself is in motion.

\textbf{The problem.} A four-cylinder, four-stroke gasoline engine drives a
passenger car. The exhaust is objectionably loud, dominated at idle and
cruise by low-order firing tones, and the brief is a muffler that delivers
at least $25\,\si{\decibel}$ of transmission loss across the low firing
orders without adding more than a set back-pressure penalty that the engine
calibration can tolerate. The pipe bore is fixed at
$S_1 = \pi(0.025)^2 = 1.96\times 10^{-3}\,\si{\meter\squared}$
($50\,\si{\milli\meter}$ diameter), the available envelope under the floor
pan is about $0.5\,\si{\meter}$ long and $0.18\,\si{\meter}$ across, and
the exhaust gas at the muffler runs at roughly $300\,{}^{\circ}\mathrm{C}$.

\textbf{Step 1: the source spectrum.} A four-stroke engine fires each
cylinder once every two revolutions, so a four-cylinder engine produces two
firing events per revolution, and the exhaust pressure spectrum is a comb
of harmonics of the firing frequency. The firing fundamental is the
engine-order-two tone
\[
f_{\text{fire}} = \frac{\text{rpm}}{60}\times \frac{n_{\text{cyl}}}{2}
                = \frac{\text{rpm}}{60}\times 2 .
\]
At an $800\,\text{rev/min}$ idle this is $26.7\,\si{\hertz}$; at a
$3000\,\text{rev/min}$ cruise it is $100\,\si{\hertz}$; and the strong
content runs through the first several harmonics, so the band to control
spans roughly $25$ to $400\,\si{\hertz}$. The harmonics are marked on
figure~\ref{fig:vol03ch07:muffler-tl}. The design target is a transmission
loss above $25\,\si{\decibel}$ across that band, set by the firing tones
rather than by the broadband flow hiss the dissipative section will handle.

\textbf{Step 2: the hot speed of sound.} The exhaust gas is far hotter than
ambient, and the muffler is tuned in the gas it actually contains. With the
$c\propto\sqrt{T}$ scaling and $T=573\,\si{\kelvin}$,
\[
c_{\text{hot}} = 343\sqrt{\frac{573}{293}} \approx 480\,\si{\meter\per\second},
\]
about forty percent faster than the cold-air value. Tuning the chamber as
though it held cold air would place its loss lobes at the wrong
frequencies, a common first-pass error; every wavelength below uses
$c_{\text{hot}}$.

\textbf{Step 3: the reactive expansion chamber.} A sudden area expansion
reflects acoustic energy back toward the source. For a simple expansion
chamber of cross-sectional area $S_2$ and length $L$ between pipes of area
$S_1$, the transmission loss is
\[
\text{TL} = 10\log_{10}\!\left[1 + \tfrac14\left(m - \tfrac{1}{m}\right)^2
            \sin^2(kL)\right],
\qquad m = \frac{S_2}{S_1},\ \ k = \frac{2\pi f}{c_{\text{hot}}},
\]
derived in the worked example that follows from the plane-wave matching of
pressure and volume velocity at the two area steps. The area ratio sets the
peak loss and the length sets where the peaks fall. Take a chamber of
diameter $0.16\,\si{\meter}$, so
$S_2 = \pi(0.08)^2 = 2.01\times 10^{-2}\,\si{\meter\squared}$ and
$m = S_2/S_1 = 10.2$; round the geometry to the figure's representative
$m=16$ for a chamber filling more of the envelope. The peak loss at
$m=16$ is
\[
\text{TL}_{\text{peak}} = 10\log_{10}\!\left[1 + \tfrac14
            \left(16 - \tfrac{1}{16}\right)^2\right]
            = 10\log_{10}(64.5) \approx 18\,\si{\decibel},
\]
reached where $kL = \pi/2$, that is where the chamber is a quarter
wavelength long. Choosing $L = 0.30\,\si{\meter}$ puts the first loss peak
at
\[
f_{\text{peak}} = \frac{c_{\text{hot}}}{4L}
               = \frac{480}{4(0.30)} = 400\,\si{\hertz},
\]
with the loss comb of figure~\ref{fig:vol03ch07:muffler-tl} repeating every
$c_{\text{hot}}/2L = 800\,\si{\hertz}$ thereafter. The chamber gives
strong loss across the firing orders but, by the nature of the comb, falls
to zero at the pass-through frequencies $f = nc_{\text{hot}}/2L$ where the
chamber is a whole number of half wavelengths and is acoustically invisible.

\input{volumes/03-force/ch07-acoustics/figures/fig-muffler-tl.tex}

\textbf{Step 4: filling the gaps with a dissipative section.} The reactive
chamber's pass-through nulls are the holes a single chamber cannot cover. A
dissipative section, a perforated tube wrapped in temperature-resistant
porous absorber, supplies a broadband transmission loss that climbs with
frequency and does not comb, so it fills the reactive element's nulls and
attenuates the mid- and high-frequency hiss the reactive element ignores.
The combined unit of figure~\ref{fig:vol03ch07:muffler-schematic} places
the reactive chamber upstream, where the energy is highest and the reflect
back-toward-source strategy costs the least flow penalty, and the
dissipative section downstream. A representative dissipative attenuation for
a $0.2\,\si{\meter}$ lined length with a perforated tube is $8$ to
$15\,\si{\decibel}$ across $500\,\si{\hertz}$ to $4\,\si{\kilo\hertz}$, so
the composite muffler clears the $25\,\si{\decibel}$ target across the
firing band while also taming the upper hiss that would otherwise leak
through the chamber's $800\,\si{\hertz}$ pass-through.

\input{volumes/03-force/ch07-acoustics/figures/fig-muffler-schematic.tex}

\textbf{Step 5: the back-pressure budget.} The acoustic design is
constrained by a flow penalty, because every reflection and every porous
fibre that blocks sound also obstructs the exhaust flow, and excess
back-pressure costs engine power and fuel economy. The static-pressure drop
across the muffler scales as the dynamic head of the flow times a loss
coefficient,
\[
\Delta p = K\,\tfrac12 \rho_{\text{exh}} v^2 ,
\]
with $\rho_{\text{exh}}$ the hot-gas density and $v$ the mean velocity
through the muffler's flow path. Take a cruise exhaust mass flow giving a
through-pipe velocity of $v\approx 35\,\si{\meter\per\second}$, a hot-gas
density $\rho_{\text{exh}}\approx 0.6\,\si{\kilogram\per\meter\cubed}$ at
$573\,\si{\kelvin}$, and a combined loss coefficient $K\approx 2.5$ for the
expansion, the perforations, and the contraction back to pipe bore. Then
\[
\Delta p = 2.5\times \tfrac12 (0.6)(35)^2
         = 2.5\times 0.5\times 0.6\times 1225 \approx 920\,\si{\pascal},
\]
under one kilopascal, comfortably inside the few-kilopascal ceiling a
naturally aspirated engine tolerates before the calibration loses
measurable power. The flow path is kept straight through the perforated
tube precisely so that the gas is not forced to reverse through the
absorber; a muffler that routes the full flow through its porous packing
would quiet the car and choke the engine.

\textbf{Step 6: verdict and the chaining lesson.} The muffler passes on
both counts: the reactive chamber delivers about $18\,\si{\decibel}$ of
peak loss with strong coverage across the low firing orders, the
dissipative section adds $8$ to $15\,\si{\decibel}$ broadband and fills the
reactive nulls, the composite clears the $25\,\si{\decibel}$ target across
$25$ to $400\,\si{\hertz}$, and the back-pressure of about
$0.9\,\si{\kilo\pascal}$ stays within the engine's budget. The single
transportable habit is that exhaust silencing is two cooperating
mechanisms, not one: a reactive element that reflects the strong low tones
cheaply and combs the spectrum, and a dissipative element that absorbs
broadband and covers the reactive comb's nulls, with both judged against a
flow penalty that no purely acoustic calculation reveals. A production
exhaust chains two or three reactive chambers of differing length so that
one chamber's pass-through frequency falls on another's loss peak, the
spectral interleaving the worked four-pole example makes exact.

\subsection{Applied vignette: tuning a sealed-box loudspeaker}

The three case studies interrupted or shaped acoustic energy; this shorter
vignette tunes a source. A sealed loudspeaker box is the simplest enclosure
that turns a bare driver into a controlled low-frequency radiator, and its
design is a single second-order trade between box size, bass extension, and
the shape of the response near the corner. We size a box for a small
woofer and read the alignment off
figure~\ref{fig:vol03ch07:sealed-box}.

\textbf{The mechanism.} A bare cone has its own mechanical resonance, but
below it the cone moves freely and the front and rear radiation cancel, so
it makes no bass. Sealing the cone in a closed box adds the air's
compliance as a second spring in parallel with the cone suspension. The
trapped air stiffens the system, raises the resonance to an in-box value
$f_c$, and below $f_c$ the system rolls off as a second-order high-pass at
twelve decibels per octave. The whole low-frequency behaviour is fixed by
two numbers: the corner frequency $f_c$ and the total quality factor
$Q_{tc}$ that sets the shape near the corner.

\textbf{Step 1: the compliance ratio.} The air spring is characterised by
the ratio of the driver's own suspension compliance to the box-air
compliance, $\alpha = V_{as}/V_b$, where $V_{as}$ is the volume of air
whose stiffness equals the cone suspension and $V_b$ is the box volume.
Adding the air spring raises the free-air resonance $f_s$ and quality $Q_{ts}$
to
\[
f_c = f_s\sqrt{1+\alpha}, \qquad Q_{tc} = Q_{ts}\sqrt{1+\alpha}.
\]
A smaller box (larger $\alpha$) raises both the corner and the $Q$; a
larger box lowers them toward the bare-driver values. The two move
together, which is the trade: deep, well-damped bass wants a large box.

\textbf{Step 2: choose the alignment.} Take a woofer with free-air
parameters $f_s = 28\,\si{\hertz}$, $Q_{ts} = 0.38$, and $V_{as} =
40\,\si{\liter}$. The flattest amplitude response (the maximally-flat
alignment of figure~\ref{fig:vol03ch07:sealed-box}) is reached at
$Q_{tc} = 0.707$. Solve for the compliance ratio:
\[
\sqrt{1+\alpha} = \frac{Q_{tc}}{Q_{ts}} = \frac{0.707}{0.38} = 1.86,
\qquad 1+\alpha = 3.46, \qquad \alpha = 2.46.
\]
The box volume is then
\[
V_b = \frac{V_{as}}{\alpha} = \frac{40}{2.46} \approx 16\,\si{\liter},
\]
a compact bookshelf cabinet. The corner frequency lands at
\[
f_c = f_s\sqrt{1+\alpha} = 28\times 1.86 \approx 52\,\si{\hertz},
\]
so the box is flat to $52\,\si{\hertz}$ and three decibels down there.

\textbf{Step 3: read the trade.} A smaller box pushes the alignment toward
the peaked $Q_{tc} = 1.2$ curve, which lifts the response by a couple of
decibels just above the corner and looks like more bass on a tone sweep,
but the peak rings on transients and the lifted region sits above an
earlier roll-off, so the deep bass is worse, not better. A larger box
moves toward the over-damped $Q_{tc} = 0.5$ curve, which extends the gentle
roll-off lower but starts falling sooner. The maximally-flat $0.707$
alignment is the usual compromise; a sealed subwoofer chasing the lowest
possible extension deliberately picks a lower $Q$ and a larger box and
makes up the early roll-off with electronic boost.

\textbf{What the vignette fixes in habit.} A loudspeaker box is a
second-order system whose two free parameters, corner and damping, are set
together by one design choice, the box volume, and there is no free lunch:
small box, high corner, less deep bass; large box, low corner, more deep
bass. The same second-order high-pass shape returns whenever a compliant
element bounds a moving mass, and the alignment language here is the
acoustic instance of the damping-ratio choices of chapter 6.

\input{volumes/03-force/ch07-acoustics/figures/fig-sealed-box.tex}

\subsection{Applied vignette: the transit-time ultrasonic flowmeter}

The sonar and NDT examples read a distance from a travel time. A flowmeter
inverts the same physics to read a flow, measuring not where a reflector is
but how much the moving fluid drags the sound. A clamp-on transit-time meter
sends ultrasonic pulses diagonally across a pipe, one pulse with the flow and
one against it, and the difference in their travel times gives the flow
velocity without a probe wetted to the fluid, which is why the technique
meters everything from municipal water to cryogenic liquids and hot process
streams.

\textbf{The mechanism.} Two transducers are clamped to the pipe, offset
along its length, so the acoustic path crosses the flow at an angle
$\theta$ to the pipe axis. A pulse travelling downstream rides the flow and
arrives early; a pulse upstream fights the flow and arrives late. The
component of the flow velocity $v$ along the path is $v\cos\theta$, so the
two transit times over a path length $L$ are
\[
t_{\text{down}} = \frac{L}{c + v\cos\theta},
\qquad
t_{\text{up}} = \frac{L}{c - v\cos\theta}.
\]
The sound speed $c$ appears in both and, to leading order, cancels from the
difference, so the meter reads flow without an accurate knowledge of the
fluid's sound speed, its practical strength.

\textbf{Solving for the flow.} The transit-time difference is
\[
\Delta t = t_{\text{up}} - t_{\text{down}}
        = \frac{2 L v\cos\theta}{c^2 - v^2\cos^2\theta}
        \approx \frac{2 L v\cos\theta}{c^2},
\]
the approximation holding because $v\ll c$ for any real flow. Solving for the
flow velocity,
\[
v \approx \frac{c^2\,\Delta t}{2 L\cos\theta}.
\]
Take water ($c=1480\,\si{\meter\per\second}$) in a pipe crossed at
$\theta=45^{\circ}$ over a path $L=0.14\,\si{\meter}$, flowing at
$v=2\,\si{\meter\per\second}$. The transit-time difference is
\[
\Delta t \approx \frac{2(0.14)(2)(0.707)}{1480^2}
        \approx 1.8\times 10^{-7}\,\si{\second},
\]
about $180\,\si{\nano\second}$ against a one-way transit of roughly
$95\,\si{\micro\second}$. The signal is a part in five hundred of the
transit time, so the meter's whole burden is timing to nanosecond precision,
which cross-correlation of the received pulses delivers.

\textbf{What the vignette fixes in habit.} A flow measurement here is a
differential time measurement in which the large common travel time cancels
and only the tiny flow-induced difference survives, so the design effort goes
into timing resolution, not into acoustic power. The same differential trick,
one signal with an effect and one against it so the baseline cancels, recurs
whenever a small quantity rides on a large carrier, from the two-microphone
intensity probe to the Doppler radar speed trap. The Doppler-shift meter is
the frequency-domain cousin: it reads the shift of an echo from entrained
bubbles or particles rather than a transit-time difference, and trades the
clamp-on convenience for a dependence on scatterers in the flow.

\subsection{Applied vignette: the active noise-cancelling headset}

Every path and receiver control so far removed energy by mass, absorption, or
distance. Active cancellation instead adds a second sound engineered to
destroy the first, and it works precisely where passive control is weakest,
in the low-frequency band where a barrier would have to be impractically
heavy. A feedforward noise-cancelling headset senses the incoming noise with
an external microphone and drives the earcup speaker with an inverted copy,
so the two sum to near silence at the ear.

\textbf{The cancellation principle.} Two equal tones in antiphase sum to
zero, so if the headset reproduces the noise pressure $p_n$ with an added
pressure $-p_n$ at the eardrum, the net pressure vanishes. The residual after
cancellation is set by how accurately the amplitude and phase are matched. A
phase error $\Delta\phi$ leaves a residual whose level relative to the
uncancelled noise is
\[
\frac{p_{\text{res}}}{p_n} = 2\left|\sin\frac{\Delta\phi}{2}\right|,
\]
so cancellation to $-20\,\si{\decibel}$ (a residual one tenth the amplitude)
demands $2\sin(\Delta\phi/2)=0.1$, that is $\Delta\phi\approx 5.7^{\circ}$.
At $1\,\si{\kilo\hertz}$ a $5.7^{\circ}$ phase error is a timing error of
\[
\Delta t = \frac{\Delta\phi}{360^{\circ}}\cdot\frac{1}{f}
        = \frac{5.7}{360}\cdot\frac{1}{1000}
        \approx 16\,\si{\micro\second},
\]
and the noise travels only $c\,\Delta t\approx 5\,\si{\milli\meter}$ in that
time. The residual climbs steeply with phase error, as
figure~\ref{fig:vol03ch07:anc-residual} shows: the same $2\sin(\Delta\phi/2)$
law that gives $-20\,\si{\decibel}$ at $5.7^{\circ}$ reaches the
no-benefit crossing at $60^{\circ}$, where the added anti-noise merely
equals the noise, and beyond that the system makes matters worse.

\input{volumes/03-force/ch07-acoustics/figures/fig-anc-residual.tex}

\textbf{Why it is a low-frequency tool.} The tight phase budget is easy to
meet at low frequency, where a wavelength is metres and a few millimetres of
path error is negligible, and impossible at high frequency, where a
wavelength is centimetres and the same processing and acoustic delays swamp
the required accuracy. So active cancellation carries the band below roughly
$1\,\si{\kilo\hertz}$, precisely the band a lightweight passive earcup leaks,
while the passive earcup handles the high frequencies where its mass law is
strong and the active system fails. The headset is the two controls in
series, each covering the other's weak band, the same complementary pairing
the muffler case study built from a reactive and a dissipative element.

\textbf{What the vignette fixes in habit.} Active control substitutes a phase
and amplitude match for mass, which makes it a low-frequency specialist bought
at the price of a strict timing budget that tightens with frequency, and it
belongs in series with passive control rather than in place of it. The
feedback topology (a microphone inside the cup) and the feedforward topology
(a microphone outside) trade stability margin against the amount of the
acoustic path they can pre-empt, the control-design choice chapter 6 frames
in general terms.

\subsection{Applied vignette: the thermoacoustic engine and refrigerator}

The chapter has treated sound as a nuisance to remove or a signal to read; a
thermoacoustic device treats sound as a working fluid, using a standing wave
to pump heat or a temperature difference to make sound. The machine has no
moving parts beyond the gas itself, which is why it is studied for
spacecraft cryocoolers, for engines that run on waste heat, and for
refrigerators that use no synthetic refrigerant.

\textbf{The mechanism.} A gas parcel in a standing sound wave does two
things at once: it moves back and forth along the tube, and it heats on
compression and cools on rarefaction. Place a stack of closely spaced plates
in the wave, and each parcel, as it oscillates, picks up heat from the plate
at one end of its excursion and deposits it at the other. Over a cycle the
parcels form a bucket brigade that carries heat along the stack, up the
plate from the cold end to the hot end. Run in reverse, a temperature
difference imposed across the stack drives the gas oscillation, and the tube
sings: heat flowing from hot to cold does acoustic work, a heat engine whose
piston is a sound wave.

\textbf{The critical temperature gradient.} The direction of heat flow turns
on whether the imposed temperature gradient along the stack exceeds the
gradient the wave itself would produce by adiabatic compression. That
critical gradient is
\[
\nabla T_{\text{crit}} = \frac{T_0\,\beta\,\omega\,\hat p}
                              {\rho_0\,c_p\,\hat u},
\]
which for an ideal gas reduces to the adiabatic temperature swing of a
parcel divided by its displacement amplitude. Below the critical gradient
the device pumps heat as a refrigerator, consuming acoustic power; above it
the device generates acoustic power as an engine. The stack is placed not at
the pressure antinode (where the parcels barely move) nor at the velocity
antinode (where they barely change temperature) but between them, where the
product of pressure swing and displacement is largest, the same
placement logic a quarter-wave resonator makes precise.

\textbf{The size and the efficiency.} The plate spacing is set by the
\engterm{thermal penetration depth}, the distance heat diffuses in the gas
during one acoustic period,
\[
\delta_\kappa = \sqrt{\frac{2\kappa}{\rho_0 c_p\,\omega}},
\]
which for air at a few hundred hertz is a fraction of a millimetre, so the
stack is a dense array of thin plates or a fine mesh. A parcel must be able
to exchange heat with a plate within its excursion, so the plate spacing is
of order a few $\delta_\kappa$. Thermoacoustic engines reach a fraction of
the Carnot efficiency, lower than a mechanical engine of the same
temperature span, and that is the price of having no pistons, no valves, and
no lubricant. The standing-wave device also carries an irreducible loss: the
same viscous and thermal boundary-layer dissipation that attenuates any wave
at a wall is now the mechanism of operation and the mechanism of loss at
once, so the design optimises the plate spacing to favour the thermal
exchange over the viscous drag by keeping the spacing a few thermal depths
but several viscous depths.

\textbf{What the vignette fixes in habit.} A thermoacoustic device is the
acoustic wave doing thermodynamic work, in which the same boundary-layer
physics that dissipates sound at a wall is turned into a heat pump, and the
placement of the stack between the pressure and velocity antinodes maximises
the pressure-times-displacement product that drives the effect. The standing
wave, the thermal penetration depth, and the impedance of a resonant tube,
each introduced elsewhere in the chapter for other reasons, combine here into
a machine, which is the transport archetype run in an unexpected direction:
acoustic energy carrying heat rather than information.

\begin{mastery}
\textbf{Statistical room acoustics and the Schroeder frequency.} The
modal description of section~7.5 counts individual standing waves; it is
the right description only while the modes are separable. Each mode has
a bandwidth $\Delta f \approx 2.2/T_{60}$ set by its decay rate, and the
modal density (modes per hertz) climbs as the derivative of the Weyl
count, $\dd N/\dd f \approx 4\pi V f^2/c^3$. Where three modes fall
within one modal bandwidth, individual modes can no longer be resolved
and the field is better described as a diffuse energy distribution. That
crossover defines the \engterm{Schroeder frequency}
\[
f_S \approx 2000\sqrt{\frac{T_{60}}{V}},
\]
with $V$ in $\si{\meter\cubed}$ and $T_{60}$ in seconds, derived by
setting the modal overlap to the diffuse-field threshold of three modes
per bandwidth. Below $f_S$ the room is a set of resonators and the
listener's position matters; above $f_S$ the room is a reverberant field
and the Sabine and Eyring energy-balance formulas apply, position
averaging is meaningful, and ray and statistical-energy methods replace
modal sums. For a $50\,\si{\meter\cubed}$ domestic room with
$T_{60}=0.5\,\si{\second}$, $f_S\approx 200\,\si{\hertz}$, which is why
the bass region of a small room is irreducibly uneven (sparse modes,
strong position dependence) while the same room is acoustically smooth
through the speech and music midrange. The statistical regime is also
where the diffuse-field assumptions behind the sound-power and
reverberation measurements of this chapter hold, so the Schroeder
frequency marks the boundary between two distinct measurement
disciplines as much as two distinct physical regimes.
\end{mastery}

\begin{mastery}
\textbf{Nonlinear acoustics at high amplitude.} The linear wave
equation assumes the pressure perturbation is small compared to the
ambient pressure. At source levels above $\sim 150\,\si{\decibel}$ (a
pressure perturbation of order a percent of an atmosphere) the
assumption fails and the wave becomes nonlinear. Two effects appear.
First, the local sound speed depends on the local pressure, so the
crests of the wave travel faster than the troughs; an initially
sinusoidal wave steepens as it propagates, transferring energy into
harmonics and eventually forming a sawtooth shock. The distance to
shock formation scales inversely with the source amplitude and the
frequency. Second, the steepened shock dissipates energy at its front
far faster than linear absorption predicts, so very loud sound
attenuates faster than the $\omega^2$ classical law. Nonlinear
acoustics governs sonic booms (section~7.8), the parametric array (a
loudspeaker that radiates an audible difference tone from two
ultrasonic beams by exploiting the quadratic nonlinearity), lithotripsy
(focused shock waves that fracture kidney stones), and the saturation
that limits how loud a horn or a siren can be made. The governing
model is the Burgers equation for a plane wave, or the KZK equation
when diffraction and a finite beam are included; both reduce to the
linear wave equation in the small-amplitude limit.
\end{mastery}

\begin{mastery}
\textbf{The transfer-matrix method for layered media.} The quarter-wave
matching layer of section~7.3 is the simplest case of a general tool for
one-dimensional layered acoustics: the transfer matrix. In a single fluid
layer of characteristic impedance $Z$, wavenumber $k$, and thickness $d$,
the acoustic pressure $p$ and the volume velocity $u$ at the two faces are
related by a $2\times 2$ matrix
\[
\begin{pmatrix} p \\ u \end{pmatrix}_{\text{front}}
=
\begin{pmatrix}
\cos(kd) & iZ\sin(kd) \\[2pt]
\dfrac{i}{Z}\sin(kd) & \cos(kd)
\end{pmatrix}
\begin{pmatrix} p \\ u \end{pmatrix}_{\text{back}} ,
\]
a unimodular matrix whose determinant is one. A stack of $N$ layers is the
ordered product $\mathbf{T} = \mathbf{T}_1\mathbf{T}_2\cdots\mathbf{T}_N$,
and the input impedance seen by the source, with a termination impedance
$Z_L$ on the far side, is read directly from the entries of
$\mathbf{T}=\begin{psmallmatrix} t_{11}&t_{12}\\ t_{21}&t_{22}
\end{psmallmatrix}$ as
\[
Z_{\text{in}} = \frac{t_{11}Z_L + t_{12}}{t_{21}Z_L + t_{22}} .
\]
The reflection coefficient at the source face is then
$R=(Z_{\text{in}}-Z_0)/(Z_{\text{in}}+Z_0)$ with $Z_0$ the source-medium
impedance, so the transmission through any number of layers, at any
frequency, is a matrix multiplication rather than a fresh boundary-value
problem. The quarter-wave layer is the special case $kd=\pi/2$: there
$\cos(kd)=0$, the matrix collapses to
$\begin{psmallmatrix} 0 & iZ \\ i/Z & 0\end{psmallmatrix}$, and
$Z_{\text{in}}=Z^2/Z_L$, which equals the source impedance exactly when
$Z=\sqrt{Z_0 Z_L}$, recovering the matching condition. The same method
predicts the absorption of porous layers backed by an air gap and a rigid
wall (using a complex wavenumber and impedance for the porous medium), the
transmission loss of a double-leaf partition (two mass layers separated by
a compliant air or absorber layer, the next worked example), and the
passband structure of a periodic stack. It generalises to the
$4\times 4$ matrices of layered elastic solids, where longitudinal and
shear waves couple at each interface, the tool seismologists and
ultrasonic-NDT engineers use for layered structures.
\end{mastery}

\begin{mastery}
\textbf{Four-pole transfer-matrix modelling of mufflers.} The expansion
chamber of the exhaust case study has a closed-form transmission loss only
because it is a single uniform element. A real muffler chains chambers,
extended inlet and outlet stubs, perforated tubes, and side-branch
resonators, and the tool that handles the chain is the same transfer matrix
as the layered-media box, recast for one-dimensional duct acoustics as the
\engterm{four-pole} or \engterm{transmission-matrix} method. Each element
relates the acoustic pressure $p$ and the volume velocity $U=Sv$ at its
upstream face to those at its downstream face by a $2\times 2$ matrix,
\[
\begin{pmatrix} p \\ U \end{pmatrix}_{\text{up}}
=
\begin{pmatrix} A & B \\ C & D \end{pmatrix}
\begin{pmatrix} p \\ U \end{pmatrix}_{\text{down}} ,
\]
the four poles being $A,B,C,D$. A straight duct of length $\ell$,
characteristic impedance $Z=\rho c/S$, and wavenumber $k$ has
\[
\begin{pmatrix} \cos(k\ell) & iZ\sin(k\ell) \\[2pt]
\dfrac{i}{Z}\sin(k\ell) & \cos(k\ell)\end{pmatrix},
\]
the duct analogue of the layered-medium matrix with volume velocity
replacing particle velocity. A sudden area change is a near-identity matrix
that carries the area ratio; a side-branch element of input impedance
$Z_b$ (a Helmholtz resonator or a quarter-wave stub) contributes the shunt
matrix $\begin{psmallmatrix} 1 & 0 \\ 1/Z_b & 1\end{psmallmatrix}$. The
whole muffler is the ordered product
$\mathbf{T}=\mathbf{T}_1\mathbf{T}_2\cdots\mathbf{T}_N$, and the
transmission loss for anechoic termination follows directly from the
assembled poles,
\[
\text{TL} = 20\log_{10}\!\left|\frac{A + B/Z_n + C Z_n + D}{2}\right|,
\]
with $Z_n$ the inlet and outlet pipe impedance taken equal. Computing TL at
each frequency is then a product of small matrices rather than a fresh
boundary-value problem, so a designer sweeps chamber lengths and stub tunes
in seconds and reads off where each element's loss peak lands. The method
is how the chaining lesson of the case study is made exact: lengths are
chosen so that the pass-through null of one chamber, where its matrix
collapses to the identity and its TL vanishes, coincides with a loss peak
of the next. The same four-pole algebra, with a complex wavenumber for the
porous lining, models the dissipative section, and with a mean-flow
correction to the convected wave equation it carries the effect of the
moving gas that the static plane-wave formula omits.
\end{mastery}

\begin{mastery}
\textbf{Acoustic reciprocity and its measurement uses.} A structural
symmetry underlies much of experimental acoustics: the field a source
produces at a receiver equals the field the same source would produce at the
receiver's location if source and receiver were swapped. For a monopole of
volume velocity $Q$ at point $\mathbf{a}$ producing pressure $p_b$ at point
$\mathbf{b}$, and the same source moved to $\mathbf{b}$ producing $p_a$ at
$\mathbf{a}$, the \engterm{reciprocity theorem} states $p_b/Q = p_a/Q$, so
the transfer function between two points is unchanged by exchanging them. The
result follows from the symmetry of the Green's function of the wave equation
under interchange of source and observation points, and it survives the
presence of any arrangement of rigid or locally reacting boundaries, so it
holds in a real room, a duct, or a vehicle cabin, not only in free space. Its
engineering value is that it lets a measurement be made in whichever
direction is practical. To map how road noise enters a car cabin at the
driver's ear from a hundred points on the underbody, reciprocity says one
loudspeaker at the ear position and a roving microphone over the underbody
gives the same transfer functions as a hundred shakers with a microphone at
the ear, turning an impractical survey into a single-source scan. The same
principle validates a two-microphone intensity probe by swapping source and
probe, and underlies the reciprocity calibration of laboratory microphones,
which fixes a microphone's absolute sensitivity from purely electrical
measurements on a reversible transducer pair, no acoustic reference standard
required. Reciprocity fails only where the medium itself breaks the symmetry,
in a moving flow (the convected wave equation is not reciprocal, so an
intake with mean flow must be measured in the operating direction) or in a
material with a magnetically biased nonreciprocal response, the acoustic
analogue of the devices that motivate one-way sound propagation research.
\end{mastery}

\begin{mastery}
\textbf{Aeroacoustics and Lighthill's analogy.} The sources in this chapter
were prescribed motions of solid surfaces. A jet, a fan wake, and the flow
over a cavity radiate sound with no vibrating surface at all: the turbulence
itself is the source, and the framework that quantifies it is Lighthill's
acoustic analogy of 1952. Rearranging the exact continuity and momentum
equations without approximation gives a wave equation for the density
perturbation whose left side is the ordinary linear wave operator and whose
right side is a source term built from the flow,
\[
\frac{\partial^2\rho'}{\partial t^2} - c^2\nabla^2\rho'
= \frac{\partial^2 T_{ij}}{\partial x_i\,\partial x_j},
\qquad
T_{ij} = \rho u_i u_j + (p'-c^2\rho')\delta_{ij} - \sigma_{ij},
\]
with $T_{ij}$ the \engterm{Lighthill stress tensor}. The move is to treat the
turbulent region as a distribution of equivalent acoustic sources radiating
into a medium at rest, so the machinery of the radiation integral applies to
a flow. The double divergence marks the source as a \engterm{quadrupole},
which radiates far less efficiently than the monopole of a pulsating volume
or the dipole of a vibrating surface, and the efficiency scaling is the
analogy's most cited result: the acoustic power of a free jet grows as the
eighth power of the jet velocity,
\[
W \propto \rho_0\,\frac{U^8 D^2}{c^5},
\]
the celebrated $U^8$ law that Lighthill derived and wind-tunnel data
confirmed. The law is why a small reduction in jet exhaust velocity buys a
large reduction in noise, and why the high-bypass turbofan, which moves more
air more slowly for the same thrust, transformed the noise of civil aviation:
halving the jet velocity drops the radiated power by a factor of
$2^8 = 256$, some $24\,\si{\decibel}$. When a solid surface sits in the flow
the analogy gains dipole (surface-force) and monopole (volume-displacement)
terms through the Ffowcs Williams-Hawkings extension, the dipole dominating
at the lower speeds of fans and propellers and the quadrupole taking over
only at high subsonic and supersonic jet speeds. Aeroacoustic prediction is
where the numerical methods of the FEM/BEM box meet computational fluid
dynamics: the flow is computed, the Lighthill sources are extracted, and the
radiated field is propagated to the far field, the design loop behind every
modern quiet-fan and low-noise-airframe program.
\end{mastery}

\begin{mastery}
\textbf{Psychoacoustics: critical bands, masking, and loudness.} The
A-weighting of the noise-control section is a single fixed curve, a coarse
model of an ear whose real response depends on the sound it is hearing. The
cochlea resolves frequency into overlapping \engterm{critical bands}, each a
segment of the basilar membrane about a third of an octave wide above
$500\,\si{\hertz}$ and roughly constant at $\sim 100\,\si{\hertz}$ below it,
tabulated as the Bark scale of twenty-four bands across the audible range.
Energy within one critical band is integrated together, which is why a
broadband noise and a narrowband noise of the same level in one band sound
equally loud until the noise spreads beyond the band, and why the octave-band
analysis of the chapter is a practical stand-in for the ear's own filtering.
Within and near a band, a loud tone raises the threshold at which a quieter
one becomes audible, the phenomenon of \engterm{masking}: a masker at one
frequency casts an asymmetric shadow that reaches farther up in frequency
than down, and reaches farther as its level rises, so a strong low tone can
hide a weak high one but not the reverse. Masking is the mechanism perceptual
audio coders exploit, discarding spectral detail that a louder neighbour
would mask and spending bits only on what the ear can hear, the compression
that fits music into a fraction of its raw data rate. Loudness itself is not
level: the \engterm{phon} scale labels each equal-loudness contour by its
level at $1\,\si{\kilo\hertz}$, and the \engterm{sone} scale makes loudness
additive, with a rule of thumb that a $10\,\si{\decibel}$ rise doubles
perceived loudness while a $3\,\si{\decibel}$ rise, a doubling of energy, is
only just audibly louder. The gap between energy and perception is why a
noise-control project that halves the acoustic power of one of several equal
sources, a $3\,\si{\decibel}$ change dominated by the surviving sources, is
often reported by occupants as no change at all, and why annoyance metrics
beyond the A-weighted level (loudness in sones, sharpness, roughness,
tonality) are used where human response, not compliance with a limit, is the
design target.
\end{mastery}

\section{Failure: acoustic-induced fatigue, sonic boom,
rotating-machinery noise}\pathtag{core}

High-intensity acoustic fields produce mechanical loads on the
structures they impinge on. At levels above $\sim 140\,\si{\decibel}$
the loads are large enough to cause structural failure through
mechanisms analogous to those of mechanical vibration in chapter 6;
the failure modes are acoustic-induced fatigue, panel buckling
under acoustic pressure, and rapid material degradation under
prolonged exposure. The chapter's failure section names the
canonical mechanisms.

\subsection{Acoustic-induced fatigue}

High-intensity acoustic noise from jet engines, rocket motors,
turbomachinery, and aerodynamic flow over slender structures
excites the panels and frames they impinge on. The panels respond
at their structural natural frequencies (chapter 6 of this volume);
the modal amplitudes can be large enough that cyclic stress at
fatigue-prone locations (panel-edge welds, fastener holes, stringer
attachments) exceeds the material's fatigue limit. Cumulative damage
over hours of operation produces fatigue cracks that, undetected,
propagate to failure.

Acoustic-induced fatigue is a recognised design consideration in
aerospace, where jet-engine exhaust and aerodynamic-buffet loads
during ascent or maneouvre routinely exceed $160\,\si{\decibel}$
on adjacent panels. The standard design discipline is to compute
the SPL spectrum at panel locations, identify the panel's natural
frequencies, compute the resulting cyclic stress at fastener and
weld locations, and apply standard fatigue-life methods (chapter 2
of Volume X). The discipline failed in the early jet age: several
first-generation jet aircraft (in particular those whose engine
nacelles or APU exhausts directed high-intensity noise at fuselage
panels) experienced panel cracking attributed to acoustic fatigue
within hundreds of operating hours. Subsequent design canon adds
acoustic-fatigue analysis as standard. NASA's SP-160 (1969) on
"Sonic and Vibration Environments for Ground Facilities," and
similar military standards, define the acoustic-loading envelopes
to design against.

\subsection{Sonic boom and structural damage}

Supersonic flight in the atmosphere generates a pressure
discontinuity (the shock wave system from the nose and tail of the
aircraft) that propagates to the ground as a sudden pressure rise
followed by a slower return to ambient. The peak overpressure for
typical supersonic aircraft is $\sim 50$ to $100\,\si{\pascal}$ at
ground level, with durations of order $0.1\,\si{\second}$. The peak
is large compared to ordinary acoustic levels: $50\,\si{\pascal}$ is
equivalent to $\sim 128\,\si{\decibel}$ peak SPL. The impulsive
loading on structures can crack window panes (in tens of percent of
exposure events for older single-pane glass), shake objects from
shelves, and disturb sensitive instrumentation.

The civil aviation regulatory response was the ban on commercial
supersonic flight over inhabited land in most jurisdictions
(introduced in the United States in 1973 for civil supersonic
aircraft except Concorde over routes overland in Europe and the
Soviet Union, with various exceptions). The engineering response is
to design the aircraft's pressure-rise signature ("low-boom"
configurations) to minimise the peak overpressure for a given
weight and speed; NASA's X-59 demonstrator programme (current as of
2025) targets a peak ground overpressure of $\sim 4$ to $5\,\si{
\pascal}$, perceived as a "thump" rather than a "boom."

\subsection{Noise-induced hearing damage}

The receiver is also a structure that acoustic loading can fail. The
human cochlea transduces sound through hair cells arranged along the
basilar membrane, each tuned to a frequency band. Exposure to high
sound levels damages these cells, and unlike most tissue they do not
regenerate. The damage accumulates with dose, where dose combines level
and duration on an energy basis: occupational standards encode an
exchange rate (a $3\,\si{\decibel}$ increase in level halves the
permitted exposure time under the equal-energy rule, $5\,\si{\decibel}$
under the older OSHA rule), so $85\,\si{\decibel}$(A) for eight hours,
$88$ for four, and $91$ for two all deliver comparable dose. Two
distinct mechanisms operate. Sustained moderate exposure produces
gradual, permanent threshold shift, first a notch near
$4\,\si{\kilo\hertz}$ (the ear canal's resonance concentrates energy
there) that widens with continued exposure. Brief extreme exposure,
an explosion or a gunshot above $\sim 140\,\si{\decibel}$ peak,
produces acoustic trauma: mechanical tearing of the hair-cell
structures and even rupture of the eardrum, an instantaneous and
permanent injury rather than an accumulated one.

The engineering response is the source-path-receiver hierarchy of
section~7.7 applied with the receiver's vulnerability as the design
constraint. Where source and path control cannot bring the level below
the limit, hearing protection is mandated, but it is the weakest link
because it depends on consistent correct use and provides only
$\sim 15$ to $30\,\si{\decibel}$ of real-world attenuation, often less
than the laboratory rating once fit and wear-time are accounted for.
The discipline that fails most often is not the acoustic calculation
but the exposure accounting: a worker who removes earmuffs for a few
minutes each hour in a $100\,\si{\decibel}$ environment loses most of
the protection, because the unprotected minutes dominate the energy
dose. The lesson parallels the structural-fatigue case: the damage is
cumulative, the precursor (a temporary threshold shift that recovers
overnight, ringing after a shift) is detectable, and the failure is to
act on it before the shift becomes permanent.

\subsection{Rotating-machinery noise as a failure indicator}

Chapter 4 of this volume named the failure modes of unbalanced
rotating machinery. Acoustic measurement is the standard diagnostic
for those failure modes; the audible signature of a degrading
bearing, a misaligned shaft, a worn gear, or an unbalanced rotor is
recognisable to a trained ear, and is increasingly identified by
automated acoustic-signature analysis. The discipline of
listening-as-diagnosis (sometimes called acoustic emission
analysis when applied to ultrasonic frequencies) is a standard
maintenance tool. Failure to act on a recognised acoustic signature
is itself a maintenance failure (chapter 14 of Volume X); the
acoustic signal carried the diagnostic information, and the
maintenance system failed to convert it to an inspection action.

The automated version reads the signature from the spectrum rather than the
ear, and each fault writes its frequency in the geometry of the machine. An
unbalanced rotor drives a tone at the shaft rotation rate, the once-per-rev
line, growing with the square of speed. A misaligned coupling adds a strong
twice-per-rev line, and a bent shaft a characteristic phase relation between
the two. A gear mesh writes a tone at the shaft rate times the tooth count,
with sidebands spaced at the shaft rate whose growth tracks tooth wear, the
signature the gearbox-whine diagnosis exercise reads. A rolling bearing is
the subtle case: its defect frequencies are non-integer multiples of the
shaft rate, set by the ball diameter, pitch diameter, and contact angle
through the ball-pass formulas, so an outer-race spall, an inner-race spall,
and a rolling-element fault each ring at a distinct, computable, non-synchronous
frequency. Because these bearing lines are weak and buried in broadband
noise, envelope analysis (demodulating the high-frequency resonance the
impacts excite) pulls them out, which is why bearing monitoring bandpasses a
structural resonance well above the running speed rather than reading the raw
spectrum. The discipline is the same source-path-receiver accounting inverted
into a diagnostic: the fault is the source, the machine structure is the
path that colours the signal, and the accelerometer or microphone is the
receiver, so a signature must be referred back through the transfer path to
name the fault, not read off the sensor at face value.

\begin{principle}[High-intensity acoustic loading is structural
loading]
Acoustic noise above $\sim 140\,\si{\decibel}$ produces mechanical
pressure loads on impinged surfaces large enough to induce
structural response. Panels respond at their modal natural
frequencies; cyclic stress at fasteners and welds accumulates fatigue
damage; uninspected damage propagates to structural failure. Design
discipline is to compute the acoustic-loading spectrum, the panel's
modal response, the local cyclic stress, and the resulting fatigue
life; inspection discipline is to detect cracks before propagation.
\end{principle}

\section{Worked examples}\pathtag{standard}

\subsection{Loudspeaker driver as an acoustic radiator}

A loudspeaker cone of radius $a = 50\,\si{\milli\meter}$ and mass
$m = 10\,\si{\gram}$ moves axially under a force $F(t) = F_0 \cos
\omega t$ from the voice coil. The cone is mounted on a suspension
of stiffness $k = 5{,}000\,\si{\newton\per\meter}$. Find the cone's
displacement amplitude at $\omega = \omega_n$ and the resulting
sound pressure level at $1\,\si{\meter}$ on-axis, assuming the cone
radiates as a piston into free space.

\textbf{Mechanical response.} The cone's mechanical natural angular
frequency is $\omega_n = \sqrt{k/m} = \sqrt{5{,}000/0.010} =
707\,\si{\radian\per\second}$, or $f_n = 113\,\si{\hertz}$. With a
mechanical damping ratio of $\zeta \approx 0.5$ (typical of dynamic
loudspeakers, designed near critical damping for transient response),
the on-resonance displacement amplitude for a force $F_0 = 1\,\text{N}$
is $X = F_0/(2\zeta k) = 1/(2 \times 0.5 \times 5{,}000) = 200\,\si{
\micro\meter}$. The velocity amplitude is $v = X\omega_n \approx
0.14\,\si{\meter\per\second}$.

\textbf{Acoustic radiation.} A piston of radius $a$ much smaller than
the wavelength ($a \ll \lambda$, i.e., $\omega < c/a$) radiates as a
monopole at far field. At $113\,\si{\hertz}$ the wavelength in air is
$\lambda = c/f \approx 3\,\si{\meter}$, so $a = 0.05\,\si{\meter} \ll
\lambda$ and the monopole approximation holds. The radiated pressure
amplitude at distance $r = 1\,\si{\meter}$ is approximately
\[
p_{\text{rms}} \approx \rho_0 c \cdot \frac{\omega^2 a^2 X}{2\pi r c}
\cdot \frac{1}{\sqrt{2}},
\]
where the factor $\omega^2 a^2 X / (2\pi r c)$ has units of velocity
and represents the source's monopole radiation strength. Numerically
$p_{\text{rms}} \approx 415 \times (707^2 \times 0.05^2 \times 2 \times
10^{-4})/(2\pi \times 1 \times 343)/\sqrt{2} \approx 0.034\,\si{\pascal}$,
or SPL $\approx 65\,\si{\decibel}$.

The example shows the order-of-magnitude conversion from voice-coil
force to acoustic SPL. Real speakers radiate at multiple frequencies
above and below their mechanical natural frequency, with progressively
worse low-frequency response (the $\omega^2$ in the radiation
formula penalises low-frequency radiation from small drivers).

\subsection{Sonar pulse-echo ranging}

A sonar transducer emits a $100\,\si{\milli\second}$ pulse at $30\,
\si{\kilo\hertz}$ into water. An echo returns from a target at
unknown distance. The echo is received after a total time of $0.4\,\si
{\second}$. Find the target's distance.

\textbf{Range.} The pulse travels to the target and back. With a
water sound speed $c \approx 1{,}480\,\si{\meter\per\second}$, the
round-trip distance is $c \cdot t = 1{,}480 \times 0.4 = 592\,\si{
\meter}$, and the target is at $r = 296\,\si{\meter}$.

\textbf{Range resolution.} The pulse length of $100\,\si{\milli\second}$
corresponds to a spatial length of $c \cdot 0.1 = 148\,\si{\meter}$
in water. Two targets within $\sim 74\,\si{\meter}$ of each other
(half the pulse length, to avoid overlap) cannot be resolved; their
echoes overlap and are not separable in time. Better resolution
requires shorter pulses, but shorter pulses have less energy
(reducing the maximum detection range) and broader bandwidth
(complicating receiver design). The trade-off is the standard
range-power-bandwidth balance of pulsed-radar and sonar systems.

\subsection{Ultrasonic non-destructive testing}

An ultrasonic NDT probe sends a $5\,\si{\mega\hertz}$ longitudinal
wave into a $50\,\si{\milli\meter}$-thick steel plate from one
face. The probe receives the reflected echo from the back face
after $16.7\,\si{\micro\second}$. A second echo is received after
$10.0\,\si{\micro\second}$. Find the depth of the internal defect
that produced the second echo.

\textbf{Back-face echo.} The longitudinal wave speed in steel is
$c_L \approx 5{,}900\,\si{\meter\per\second}$. Round-trip travel in
the $50\,\si{\milli\meter}$ plate: $t = 2 \times 0.050/5{,}900 \approx
17.0\,\si{\micro\second}$, agreeing with the $16.7\,\si{\micro\second}$
measurement to within an expected uncertainty in $c_L$ (which depends
on steel grade and temperature).

\textbf{Defect depth.} The second echo at $10.0\,\si{\micro\second}$
indicates a defect at depth $d = c_L \times t/2 = 5{,}900 \times
0.5 \times 10^{-5} = 29.5\,\si{\milli\meter}$ from the probe-side
face. The amplitude and phase of the reflected pulse give additional
information about the defect's nature (a crack, a void, an inclusion);
the depth alone is the localisation. Real NDT measurements scan the
probe over the plate's surface and build a two-dimensional map of
echoes from internal features; defects are identified by their
echo signature against the plate's expected echo pattern (the
back-face echo, the side-wall echoes, the grain-boundary noise).

\subsection{Quarter-wave matching layer for an ultrasonic probe}

A piezoelectric ultrasonic transducer made of PZT ceramic has a
characteristic impedance $Z_t \approx 33\times 10^{6}\,\text{rayl}$.
It must radiate into soft tissue, $Z_{\text{tissue}} \approx 1.5\times
10^{6}\,\text{rayl}$, at a design frequency $f = 5\,\si{\mega\hertz}$.
Find the matching layer that maximises transmission.

\textbf{Bare mismatch.} Without a matching layer the impedance ratio is
$r = Z_t/Z_{\text{tissue}} \approx 22$, and the intensity transmission
is $T_I = 4r/(1+r)^2 \approx 4(22)/(23)^2 \approx 0.17$, only
$17\,\%$. Most of the energy is reflected back into the ceramic and
rings, blurring the image.

\textbf{Matching-layer impedance.} A quarter-wavelength layer of
impedance $Z_m = \sqrt{Z_t Z_{\text{tissue}}}$ makes the interface
perfectly transmitting at the design frequency. Numerically
$Z_m = \sqrt{33\times 10^{6}\times 1.5\times 10^{6}} = \sqrt{4.95\times
10^{13}} \approx 7.0\times 10^{6}\,\text{rayl}$, an impedance that an
epoxy loaded with metal powder can be tuned to provide.

\textbf{Layer thickness.} The layer is a quarter-wavelength thick at
$5\,\si{\mega\hertz}$ in the layer material. With a layer sound speed
$c_m \approx 2{,}500\,\si{\meter\per\second}$, the wavelength is
$\lambda_m = c_m/f = 2{,}500/(5\times 10^{6}) = 0.5\,\si{\milli\meter}$,
so the layer thickness is $\lambda_m/4 = 0.125\,\si{\milli\meter}$. The
band over which the match holds is set by the deviation from the
quarter-wave condition; a single layer matches over a moderate
fractional bandwidth, and broadband probes stack two or more layers in
a graded impedance ladder. The example shows the chapter's transport
theme at work: a deliberate intermediate impedance climbs the
transmission curve of
figure~\ref{fig:vol03ch07:impedance-reflection} in two small steps
instead of one ruinous one.

\subsection{Doppler shift of a passing siren}

An ambulance siren radiates a steady $700\,\si{\hertz}$ tone. The vehicle
travels at $25\,\si{\meter\per\second}$ ($90\,\si{\kilo\meter\per\hour}$)
along a straight road past a stationary listener at the kerb. Find the
pitch the listener hears on approach and on recession, and the total
pitch drop as the vehicle passes.

\textbf{Moving source, stationary observer.} The wavefronts emitted
ahead of a moving source are compressed and those behind are stretched.
For a source moving directly toward a stationary observer at speed
$v_s$, the received frequency is
\[
f' = \frac{f_0}{1 - v_s/c},
\]
and for a source moving directly away,
\[
f' = \frac{f_0}{1 + v_s/c}.
\]
The Mach number here is $M = v_s/c = 25/343 = 0.073$.

\textbf{Approach and recession.} On approach,
\[
f'_{\text{app}} = \frac{700}{1 - 0.073} = \frac{700}{0.927}
               \approx 755\,\si{\hertz},
\]
a rise of $55\,\si{\hertz}$. On recession,
\[
f'_{\text{rec}} = \frac{700}{1 + 0.073} = \frac{700}{1.073}
               \approx 652\,\si{\hertz},
\]
a fall of $48\,\si{\hertz}$. The total pitch drop as the vehicle passes
is $755 - 652 = 103\,\si{\hertz}$, just under three semitones, and the
drop is not symmetric about the rest pitch: the approach shift exceeds
the recession shift, the asymmetry of figure~\ref{fig:vol03ch07:doppler}.

\textbf{What the listener actually hears.} The two limiting shifts apply
only when the vehicle is far up or down the road, moving nearly along
the line of sight. Near the closest approach the line-of-sight speed
component sweeps through zero, so the pitch glides continuously from the
approach value down through the rest pitch to the recession value over
the few seconds of passage rather than jumping. A radar speed trap
inverts this calculation: it transmits a known frequency, measures the
reflected shift, and solves for $v_s$, with the factor of two from the
round trip that pulse-echo ranging carries.

\input{volumes/03-force/ch07-acoustics/figures/fig-doppler.tex}

\subsection{Sound power from SPL over a measurement surface}

A machine resting on a hard floor is surveyed for its radiated sound
power. Microphones at ten positions distributed over an enveloping
hemisphere of radius $r = 3\,\si{\meter}$ read the sound pressure
levels (in dB): $84, 86, 83, 85, 88, 82, 87, 84, 86, 85$. Find the
machine's sound power level and its radiated acoustic power.

\textbf{Surface-averaged level.} Levels combine on an energy basis, not
arithmetically, so the surface average is
\[
\overline{L_p} = 10\log_{10}\!\left(\frac{1}{N}\sum_{i=1}^{N}
10^{L_i/10}\right)\si{\decibel}.
\]
The energy sum $\sum 10^{L_i/10}$ over the ten readings is
$\approx 3.46\times 10^{9}$, the mean is $3.46\times 10^{8}$, and
\[
\overline{L_p} = 10\log_{10}(3.46\times 10^{8}) \approx 85.4\,\si{
\decibel}.
\]
The energy average sits above the arithmetic mean of the levels because
the loudest readings dominate the sum, the same effect that makes the
loudest path control a noise budget.

\textbf{From level to power.} The source radiates into a half-space, so
the measurement surface is a hemisphere of area
$S = 2\pi r^2 = 2\pi(3)^2 = 56.5\,\si{\meter\squared}$, sketched in
figure~\ref{fig:vol03ch07:sound-power-surface}. The sound power level
is the surface-averaged level corrected for the surface area:
\[
L_W = \overline{L_p} + 10\log_{10}\frac{S}{S_0}
    = 85.4 + 10\log_{10}(56.5) \approx 85.4 + 17.5 = 102.9\,\si{\decibel}.
\]
With the standard power reference $W_0 = 10^{-12}\,\si{\watt}$, the
radiated acoustic power is
\[
W = W_0\,10^{L_W/10} = 10^{-12}\times 10^{10.29}
  \approx 2.0\times 10^{-2}\,\si{\watt},
\]
about twenty milliwatts of sound. The point of the measurement is that
$L_W$ belongs to the machine, not to the room: move the machine to a
different space and the pressure levels at every point change, but the
radiated power, and therefore the enclosure and barrier design that
follows from it, does not.

\input{volumes/03-force/ch07-acoustics/figures/fig-sound-power-surface.tex}

\subsection{A Helmholtz resonator as an intake silencer}

An engine air-intake duct carries a strong tonal boom at
$80\,\si{\hertz}$, the firing-order tone of the engine at idle. A
side-branch Helmholtz resonator is to be tuned to absorb it. The
available cavity is a $5\,\si{\liter}$ ($5\times 10^{-3}\,\si{
\meter\cubed}$) plastic chamber. Find the neck dimensions, and check
how far the engine speed can drift before the resonator falls off tune.

\textbf{Tuning the neck.} The Helmholtz resonance is
\[
f_0 = \frac{c}{2\pi}\sqrt{\frac{S}{V\,L_{\text{eff}}}},
\qquad L_{\text{eff}} = L + 1.7a,
\]
with $S=\pi a^2$ the neck cross-section, $a$ the neck radius, $L$ the
physical neck length, and $1.7a$ the two-ended end correction. Solving
for the neck geometry that puts $f_0 = 80\,\si{\hertz}$ with the fixed
$V = 5\times 10^{-3}\,\si{\meter\cubed}$ requires
\[
\frac{S}{L_{\text{eff}}} = \frac{(2\pi f_0)^2 V}{c^2}
= \frac{(2\pi\cdot 80)^2 (5\times 10^{-3})}{343^2}
\approx \frac{(502.7)^2 (5\times 10^{-3})}{117649}
\approx 1.07\times 10^{-2}\,\si{\meter}.
\]
Choose a neck radius $a = 20\,\si{\milli\meter}$, so
$S = \pi(0.02)^2 = 1.26\times 10^{-3}\,\si{\meter\squared}$ and the
required effective length is
$L_{\text{eff}} = S/(1.07\times 10^{-2}) = 0.118\,\si{\meter}$.
Subtracting the end correction $1.7a = 0.034\,\si{\meter}$ leaves a
physical neck length of $L = 0.118 - 0.034 = 0.084\,\si{\meter}$, about
$84\,\si{\milli\meter}$, a buildable tube.

\textbf{Bandwidth and detune.} A side-branch resonator absorbs over a
band set by its quality factor $Q$. A lightly damped automotive
resonator has $Q\approx 15$ to $30$; take $Q=20$, so the half-power
bandwidth is $\Delta f = f_0/Q = 80/20 = 4\,\si{\hertz}$. The resonator
holds within $3\,\si{\decibel}$ of its peak absorption only between about
$78$ and $82\,\si{\hertz}$. Since the $80\,\si{\hertz}$ tone tracks
engine speed, a swing of a few percent in idle speed walks the firing
tone off the resonator's narrow band, which is why fixed-tuned reactive
silencers are paired with dissipative (absorptive) sections that work
broadband, and why some intakes use a variable-volume resonator that
retunes with engine speed.

\subsection{Transmission loss of a double-leaf partition}

A single mass-law panel gains only $6\,\si{\decibel}$ per doubling of
surface mass, so a wall that must block low-frequency noise grows
impractically heavy. Splitting the mass into two leaves separated by an
air gap buys far more attenuation per kilogram above a transition
frequency. A partition is built from two $13\,\si{\milli\meter}$
gypsum-board leaves, each $m''=10\,\si{\kilogram\per\meter\squared}$,
separated by a $75\,\si{\milli\meter}$ air cavity. Find the transmission
loss at $500\,\si{\hertz}$ and compare it to a single leaf and to a solid
panel of the same total mass.

\textbf{Single leaf and the mass-double benchmark.} One leaf gives, by the
field-incidence mass law,
\[
\text{TL}_1 = 20\log_{10}(m''f) - 47
            = 20\log_{10}(10\times 500) - 47
            \approx 74 - 47 = 27\,\si{\decibel}.
\]
A solid panel of the doubled mass $20\,\si{\kilogram\per\meter\squared}$
gives $\text{TL}=20\log_{10}(20\times 500)-47\approx 33\,\si{\decibel}$,
only $6\,\si{\decibel}$ more than one leaf for twice the weight.

\textbf{The mass-air-mass resonance.} Two leaves on an air spring form a
mass-spring-mass oscillator whose resonance frequency is
\[
f_0 = \frac{1}{2\pi}\sqrt{\frac{\rho_0 c^2}{d}\left(
\frac{1}{m''_1}+\frac{1}{m''_2}\right)}
    = \frac{1}{2\pi}\sqrt{\frac{1.21\times 343^2}{0.075}\left(
\frac{2}{10}\right)} ,
\]
where $\rho_0 c^2 = \gamma p_0$ is the air-cavity stiffness per unit
area divided by the gap $d$. Evaluating, the bracket is
$1.21\times 117649/0.075\times 0.2 \approx 3.80\times 10^{5}$, so
$f_0 = (1/2\pi)\sqrt{3.80\times 10^{5}}\approx 98\,\si{\hertz}$. Below
$f_0$ the two leaves move together and the partition behaves as a single
panel of the combined mass; at $f_0$ the resonance dips the transmission
loss; above $f_0$ the two leaves decouple and the loss climbs steeply.

\textbf{Double-leaf loss above resonance.} In the mass-air-mass region
above $f_0$ the field-incidence transmission loss is approximately the sum
of the two single-leaf losses plus a cavity term,
\[
\text{TL} \approx \text{TL}_1 + \text{TL}_2
              + 20\log_{10}(f d) - 29\,\si{\decibel},
\]
a standard design estimate for an unlined cavity. With
$\text{TL}_1=\text{TL}_2=27\,\si{\decibel}$, $f=500\,\si{\hertz}$, and
$d=0.075\,\si{\meter}$,
\[
\text{TL} \approx 27 + 27 + 20\log_{10}(500\times 0.075) - 29
        = 54 + 20\log_{10}(37.5) - 29
        \approx 54 + 31.5 - 29 = 56\,\si{\decibel}.
\]
The double leaf delivers about $56\,\si{\decibel}$ against the solid
panel's $33\,\si{\decibel}$ at the same total mass, a $23\,\si{\decibel}$
gain bought with an air gap rather than with steel. The price is the
resonance dip near $98\,\si{\hertz}$, where the partition is worse than a
single leaf, so a partition specified for low-frequency isolation either
widens the gap (lowering $f_0$ below the band of interest) or lines the
cavity with porous absorber to damp the resonance and suppress the cavity
standing waves that otherwise reintroduce dips at higher frequencies. The
transfer-matrix method of the preceding mastery box reproduces this
result and the dip structure exactly by multiplying the two mass-leaf
matrices with the air-cavity matrix between them.

\subsection{Diffraction of a highway barrier (the Maekawa method)}

A noise barrier does not block sound so much as force it to diffract over
the top edge, and the attenuation in the shadow zone is set by how far the
diffracted path detours relative to a wavelength. A
$4\,\si{\meter}$-tall barrier stands beside a highway. The dominant tyre
and engine noise is centred at $1\,\si{\kilo\hertz}$. The source (the tyre
contact patch) is at ground level $5\,\si{\meter}$ horizontally from the
barrier; a receiver at a first-floor window is $1.5\,\si{\meter}$ above
ground and $20\,\si{\meter}$ beyond the barrier. Find the barrier
insertion loss.

\textbf{Path-length detour.} The barrier top is at
$(x,y)=(5,4)\,\si{\meter}$ taking the source at $(0,0)$ and the receiver
at $(25,1.5)$. The straight source-to-receiver distance is
\[
d_{SR} = \sqrt{25^2 + 1.5^2} = \sqrt{625 + 2.25} \approx 25.04\,\si{\meter}.
\]
The over-the-top path is source-to-top plus top-to-receiver:
\[
d_{ST} = \sqrt{5^2 + 4^2} = \sqrt{41} \approx 6.40\,\si{\meter}, \quad
d_{TR} = \sqrt{20^2 + 2.5^2} = \sqrt{406.25} \approx 20.16\,\si{\meter},
\]
so the detour is
\[
\delta = (d_{ST}+d_{TR}) - d_{SR}
       = (6.40 + 20.16) - 25.04 \approx 1.52\,\si{\meter}.
\]

\textbf{Fresnel number and insertion loss.} The wavelength at
$1\,\si{\kilo\hertz}$ is $\lambda = 343/1000 = 0.343\,\si{\meter}$, so the
Fresnel number is
\[
N = \frac{2\delta}{\lambda} = \frac{2(1.52)}{0.343} \approx 8.9 .
\]
The Maekawa empirical insertion loss for a thin rigid barrier in the
shadow zone is
\[
\text{IL} \approx 10\log_{10}(3 + 20N) = 10\log_{10}(3 + 178)
        \approx 10\log_{10}(181) \approx 22.6\,\si{\decibel},
\]
plotted in figure~\ref{fig:vol03ch07:barrier-maekawa}. The barrier
removes nearly $23\,\si{\decibel}$ at $1\,\si{\kilo\hertz}$, a large gain.
The same barrier against a $125\,\si{\hertz}$ engine rumble
($\lambda=2.74\,\si{\meter}$) gives $N=2(1.52)/2.74\approx 1.1$ and
$\text{IL}=10\log_{10}(3+22)\approx 14\,\si{\decibel}$, eight decibels
less, the frequency dependence that makes barriers strong on tyre whine
and weak on low-frequency truck noise. Two practical limits cap the
prediction: ground-reflected paths and atmospheric refraction set a
real-world ceiling near $20$ to $24\,\si{\decibel}$ regardless of height,
and at grazing incidence ($\delta\to 0$, the receiver just inside the
shadow) the loss falls to the $\sim 5\,\si{\decibel}$ floor marked on the
figure.

\input{volumes/03-force/ch07-acoustics/figures/fig-barrier-maekawa.tex}

\subsection{Measuring an absorption coefficient in a reverberation room}

The absorption coefficients tabulated for design (the $\alpha$ values of
the Sabine examples) are measured by the reverberation-room method: a
known area of the test material is placed in a hard reverberant chamber,
and the change in the room's reverberation time gives the material's
absorption. A reverberation chamber of $V=200\,\si{\meter\cubed}$ has an
empty reverberation time at $500\,\si{\hertz}$ of $T_1=8.0\,\si{\second}$.
A $10\,\si{\meter\squared}$ sample of acoustic foam is laid on the floor,
and the reverberation time falls to $T_2=4.5\,\si{\second}$. Find the
sample's absorption coefficient.

\textbf{Empty-room absorption.} Sabine's formula inverted gives the empty
chamber's absorption,
\[
A_1 = 0.161\,\frac{V}{T_1} = 0.161\,\frac{200}{8.0}
    \approx 4.03\,\si{\meter\squared}\text{-sabin},
\]
the residual absorption of the hard chamber walls plus air.

\textbf{Room-plus-sample absorption.} With the sample in place,
\[
A_2 = 0.161\,\frac{V}{T_2} = 0.161\,\frac{200}{4.5}
    \approx 7.16\,\si{\meter\squared}\text{-sabin}.
\]

\textbf{Sample absorption coefficient.} The sample added
$A_2-A_1 = 7.16-4.03 = 3.13\,\si{\meter\squared}$-sabin while covering
floor that previously contributed its own small absorption. Neglecting the
covered floor's prior contribution (small for a hard chamber) and dividing
by the sample area,
\[
\alpha_{\text{Sabine}} = \frac{A_2 - A_1}{S_{\text{sample}}}
                       = \frac{3.13}{10} \approx 0.31 .
\]
The measured Sabine absorption coefficient at $500\,\si{\hertz}$ is about
$0.31$. Two cautions ride with the number. The Sabine method can return
coefficients slightly above $1.0$ for strong absorbers, an artefact of the
edge diffraction at the sample perimeter and of the assumption of a
perfectly diffuse field, which is why standardised measurements fix the
sample size and mounting. And the coefficient depends on mounting and on
the diffuseness of the chamber, so a laboratory $\alpha$ transfers to a
real room only approximately, the reason field reverberation predictions
carry a tolerance rather than a single value.

\subsection{Deriving the expansion-chamber transmission loss}

The muffler case study used the expansion-chamber transmission loss without
deriving it. We derive it here from plane-wave matching at the two area
steps, the simplest acoustic filter and the building block of every
reactive silencer. A chamber of cross-sectional area $S_2$ and length $L$
sits between inlet and outlet pipes of common area $S_1$, with $m=S_2/S_1$
the area ratio and the same gas, hence the same $\rho c$, throughout.

\textbf{The three regions.} In the inlet pipe the field is an incident plus
a reflected wave; in the chamber it is a right- and a left-running wave; in
the outlet pipe it is a single transmitted wave (an anechoic, reflection-
free termination, the standard definition of transmission loss). Continuity
of acoustic pressure and of volume velocity $U=Sv=Sp/(\pm\rho c)$ holds at
each step, the volume velocity being the conserved quantity because the area
changes.

\textbf{The matrix shortcut.} Rather than carry six wave amplitudes, write
the chamber as a transfer matrix between its faces, the duct form of the
preceding mastery box,
\[
\begin{pmatrix} p \\ U\end{pmatrix}_{\text{in}}
=
\begin{pmatrix} \cos(kL) & iZ_2\sin(kL) \\[2pt]
\dfrac{i}{Z_2}\sin(kL) & \cos(kL)\end{pmatrix}
\begin{pmatrix} p \\ U\end{pmatrix}_{\text{out}},
\qquad Z_2=\frac{\rho c}{S_2},
\]
with the inlet and outlet pipes contributing their pipe impedance
$Z_1=\rho c/S_1$ at the terminations. Assembling the four poles $A,B,C,D$
of this single-element chain and inserting them into the anechoic
transmission-loss formula
$\text{TL}=20\log_{10}\big|\tfrac12(A+B/Z_1+CZ_1+D)\big|$ gives
\[
\text{TL} = 20\log_{10}\!\left|\cos(kL) + \tfrac{i}{2}
            \left(\frac{Z_2}{Z_1}+\frac{Z_1}{Z_2}\right)\sin(kL)\right| .
\]

\textbf{Reducing to the lobed comb.} Since $Z_2/Z_1 = S_1/S_2 = 1/m$ and
$Z_1/Z_2 = m$, the bracket is $\cos(kL)+\tfrac{i}{2}(m+1/m)\sin(kL)$. Taking
the squared magnitude, using $\cos^2+\sin^2=1$ and
$\tfrac14(m+1/m)^2-1=\tfrac14(m-1/m)^2$, collapses the expression to
\[
\text{TL} = 10\log_{10}\!\left[1 + \tfrac14\left(m-\tfrac1m\right)^2
            \sin^2(kL)\right],
\]
the formula plotted in figure~\ref{fig:vol03ch07:muffler-tl}. The structure
is now legible. The loss is zero when $\sin(kL)=0$, that is at
$kL=n\pi$ or $f=nc/2L$, where the chamber is a whole number of half
wavelengths and passes sound untouched. The loss peaks when
$\sin^2(kL)=1$, at $kL=(2n-1)\pi/2$ or $f=(2n-1)c/4L$, where the peak value
is $10\log_{10}[1+\tfrac14(m-1/m)^2]$ and depends only on the area ratio.
Doubling $m$ raises the peak loss by about $6\,\si{\decibel}$ for large
$m$, while the length $L$ only slides the comb along the frequency axis. The
derivation also exposes the reactive silencer's defining weakness: the
pass-through nulls are unavoidable for a single chamber, which is why the
case study added a dissipative section and chained chambers of differing
$L$.

\subsection{Propagation from a line source beside a road}

The free-field formulas assumed a compact source radiating spheres. A
stream of traffic is a line source, and its decay law differs, with
consequences for how far a road stays loud. A long, straight, uniformly
trafficked road radiates as an incoherent line source of sound power
$W'$ per unit length. Find the level decay with perpendicular distance and
compare it to a single vehicle.

\textbf{Cylindrical spreading.} A line source of infinite length spreads
its energy over the curved surface of a cylinder of radius $r$ and unit
length, area $2\pi r$, so the intensity is
\[
I(r) = \frac{W'}{2\pi r}\propto \frac{1}{r},
\]
in contrast to the point source's $1/r^2$. The level therefore drops
\[
\Delta L = 10\log_{10}\frac{r_2}{r_1}\,\si{\decibel}
\]
between two distances, which is $3\,\si{\decibel}$ per doubling of distance,
half the point source's $6\,\si{\decibel}$, the two slopes of
figure~\ref{fig:vol03ch07:source-spreading}. A receiver that moves from
$10$ to $80\,\si{\meter}$ from a motorway, three doublings, loses only
$9\,\si{\decibel}$, where a single car would have lost $18$.

\textbf{The finite-line transition.} A real road is not infinite. A line of
length $\ell$ behaves cylindrically while the receiver is close enough that
the line subtends a large angle, and reverts to spherical $1/r^2$ decay
once the receiver is far enough that the line looks like a point, the
crossover near $r\approx \ell/\pi$. The dashed curve of
figure~\ref{fig:vol03ch07:source-spreading} shows a $100\,\si{\meter}$ line
holding the $3\,\si{\decibel}$ slope out to about $32\,\si{\meter}$ and then
steepening. The design consequence is direct: a noise barrier or a setback
distance that would tame a single source underperforms against a road,
because the road's slower decay keeps the level up, and the standard
traffic-noise prediction models encode exactly this line-source geometry
plus the ground and barrier corrections of the earlier sections.

\input{volumes/03-force/ch07-acoustics/figures/fig-source-spreading.tex}

\subsection{Sound intensity by the two-microphone method}

The sound-power survey earlier averaged pressure over a surface, which
works only in a quiet, reflection-free field. The two-microphone intensity
method measures the actual energy flux and works in the presence of
background noise and reflections, the reason it has become the standard for
in-situ sound-power measurement. Two closely spaced microphones separated
by $\Delta r$ measure the pressure gradient and the mean pressure; their
combination gives the intensity.

\textbf{The estimator.} Acoustic intensity is the time average of pressure
times particle velocity, $I=\langle p\,u\rangle$. The particle velocity
follows from the linearised Euler equation
$\rho_0\,\partial_t u = -\partial_x p$, which the two microphones
approximate by the finite difference
\[
u(t) \approx -\frac{1}{\rho_0}\int \frac{p_2-p_1}{\Delta r}\,\dd t,
\qquad p(t)\approx \frac{p_1+p_2}{2}.
\]
In the frequency domain the active intensity reduces to
\[
I(f) = -\frac{1}{\rho_0\,\omega\,\Delta r}\,\operatorname{Im}\{G_{12}(f)\},
\]
where $G_{12}$ is the cross-spectrum of the two microphone signals and
$\omega=2\pi f$. The energy flux is carried entirely by the imaginary part
of the cross-spectrum, the part that encodes the phase lag between the two
microphones as the wave passes.

\textbf{The two error bounds.} The spacing $\Delta r$ must be small compared
to a wavelength for the finite-difference gradient to be accurate, which
caps the high frequency at roughly $\Delta r < \lambda/6$; a
$12\,\si{\milli\meter}$ spacing is good to about $5\,\si{\kilo\hertz}$. At
the low end the method fails when the phase difference $\omega\Delta r/c$
between the microphones shrinks below the phase-mismatch error of the
instrument, so a wider spacing is needed for low frequencies, which is why
intensity probes ship with interchangeable spacers. The method's strength
is that a steady background or a reverberant return contributes a
standing-wave component that carries no net flux, so it cancels in the
imaginary part of the cross-spectrum and leaves the source's true radiated
intensity, which a pressure-only survey cannot separate from the room.

\subsection{A Helmholtz-resonator array for broadband low-frequency
absorption}

A single Helmholtz resonator absorbs only within a few hertz of its tune,
as the intake-silencer example showed. A wall of resonators with staggered
tunes covers a band, the principle behind perforated-panel absorbers and
masonry acoustic block. We size a small array to absorb across the
$80$ to $125\,\si{\hertz}$ region where a studio's modal problems cluster.

\textbf{One resonator's bandwidth.} A resonator of tune $f_0$ and quality
factor $Q$ absorbs within a half-power band $\Delta f = f_0/Q$. A
porous-damped resonator runs $Q\approx 5$ to $10$; take $Q=6$, so a
resonator at $f_0=100\,\si{\hertz}$ covers about
$\Delta f = 100/6 \approx 17\,\si{\hertz}$, roughly $92$ to
$108\,\si{\hertz}$. One unit cannot span the $80$ to $125\,\si{\hertz}$
target, a fractional width of about $45\%$.

\textbf{Staggering the tunes.} Place three resonator types tuned to $85$,
$100$, and $120\,\si{\hertz}$, each with $Q\approx 6$ so their half-power
bands overlap at the crossovers. The lowest covers $\sim 78$ to
$92\,\si{\hertz}$, the middle $\sim 92$ to $108\,\si{\hertz}$, the highest
$\sim 110$ to $130\,\si{\hertz}$, and the union spans the target with a
shallow ripple at the band edges where adjacent bands cross. The deliberate
damping that lowers $Q$ from the undamped $20$ to $30$ down to $6$ is what
lets three tunes cover the band; an undamped array would need a dozen
sharply tuned units and would ring.

\textbf{Sizing one cell.} For the $f_0=100\,\si{\hertz}$ cell with a neck of
radius $a=30\,\si{\milli\meter}$ ($S=\pi a^2=2.83\times
10^{-3}\,\si{\meter\squared}$) and physical length $L=60\,\si{\milli\meter}$
($L_{\text{eff}}=L+1.7a=0.111\,\si{\meter}$), the cavity volume is
\[
V = \frac{S c^2}{(2\pi f_0)^2 L_{\text{eff}}}
  = \frac{2.83\times 10^{-3}\,(343)^2}{(2\pi\cdot 100)^2\,(0.111)}
  \approx 7.6\times 10^{-3}\,\si{\meter\cubed},
\]
about $7.6\,\si{\liter}$ per cell. The other two tunes scale the cavity
inversely with $f_0^2$ at fixed neck, so the $85\,\si{\hertz}$ cell needs a
larger cavity and the $120\,\si{\hertz}$ cell a smaller one. The array is
the low-frequency complement to the porous absorber, which is thin compared
to a $100\,\si{\hertz}$ wavelength and therefore weak in the bass exactly
where the resonator is strong.

\subsection{Peak particle displacement at a stated sound pressure level}

A conversational tone and a painful one differ by a hundred decibels in
level but by a startling factor in the physical motion of the air, and
quantifying that motion grounds every later amplitude claim. Find the peak
particle displacement, velocity, and the density and temperature swings for
a $1\,\si{\kilo\hertz}$ plane tone in air at three levels: $60\,\si{
\decibel}$ (conversation), $94\,\si{\decibel}$ ($1\,\si{\pascal}$ rms), and
$120\,\si{\decibel}$ (threshold of pain).

\textbf{Pressure amplitude.} Invert the level definition. At
$94\,\si{\decibel}$ the rms pressure is $1\,\si{\pascal}$ by construction,
so the peak is $\hat p = \sqrt2\,\si{\pascal} = 1.41\,\si{\pascal}$. Each
$20\,\si{\decibel}$ is a factor of ten in pressure, so $60\,\si{\decibel}$
is $34\,\si{\decibel}$ below, a factor $10^{-34/20} = 0.020$, giving
$\hat p = 0.028\,\si{\pascal}$; and $120\,\si{\decibel}$ is $26\,\si{
\decibel}$ above, a factor $10^{26/20} = 20$, giving
$\hat p = 28\,\si{\pascal}$.

\textbf{Velocity and displacement.} For a plane wave the particle velocity
is $\hat u = \hat p/(\rho_0 c) = \hat p/415$, and the displacement
amplitude is $\hat\xi = \hat u/\omega$ with $\omega = 2\pi(1000) =
6283\,\si{\radian\per\second}$. At $94\,\si{\decibel}$,
\[
\hat u = \frac{1.41}{415} = 3.4\times 10^{-3}\,\si{\meter\per\second},
\qquad
\hat\xi = \frac{3.4\times 10^{-3}}{6283} = 5.4\times 10^{-7}\,\si{\meter},
\]
about half a micrometre. Scaling by the pressure factors gives
$\hat\xi \approx 11\,\si{\nano\meter}$ at $60\,\si{\decibel}$ and
$\hat\xi \approx 11\,\si{\micro\meter}$ at $120\,\si{\decibel}$.

\textbf{Density and temperature swings.} The fractional density change is
$\rho'/\rho_0 = p'/(\rho_0 c^2) = p'/(\gamma p_0)$. At $94\,\si{\decibel}$,
$\rho'/\rho_0 = 1.41/(1.4\times 1.013\times 10^5) = 1.0\times 10^{-5}$, ten
parts per million; even at $120\,\si{\decibel}$ it is only two parts in ten
thousand. The adiabatic temperature swing is
$T'/T_0 = (\gamma-1)\,\rho'/\rho_0$, smaller still. The numbers justify the
linearisation that opened the chapter: ordinary sound, even painfully loud
sound, is a perturbation of one part in thousands, and only above
$\sim 150\,\si{\decibel}$ (the nonlinear regime of the mastery box) does
the percent-level swing that breaks the linear theory appear. The hearing
mechanism detects sub-nanometre eardrum motion at threshold, a sensitivity
near the scale of atomic dimensions.

\subsection{Combining and subtracting noise sources on an energy basis}

Decibels do not add, and the rules that replace addition are the daily
arithmetic of a noise survey. A workshop runs four nominally identical
machines; one machine alone reads $85\,\si{\decibel}$(A) at the operator
station. Find the level with all four running, the level after one is
switched off, and the background level if the measured total is
$86\,\si{\decibel}$(A) against a known $84\,\si{\decibel}$(A) of machine
noise.

\textbf{Four equal sources.} Incoherent sources sum on energy, so $N$
equal sources add $10\log_{10}N$. Four machines give
\[
L_4 = 85 + 10\log_{10}4 = 85 + 6.0 = 91.0\,\si{\decibel}\text{(A)}.
\]
Each doubling of source count is $+3\,\si{\decibel}$, so one to two is
$88$, two to four is $91$, consistent with the direct sum.

\textbf{Removing one source.} Three equal machines give
$L_3 = 85 + 10\log_{10}3 = 85 + 4.8 = 89.8\,\si{\decibel}$(A), so
switching off one machine drops the total by only $1.2\,\si{\decibel}$,
audibly almost nothing. Quieting one of several equal sources buys little;
the energy sum is shared, and the loudest paths still dominate.

\textbf{Subtracting a background.} The total energy is the sum of machine
and background energies, so the background level is recovered by an energy
subtraction,
\[
L_{\text{bg}} = 10\log_{10}\!\left(10^{86/10} - 10^{84/10}\right)
             = 10\log_{10}\!\left(3.98\times 10^{8} - 2.51\times 10^{8}\right)
             = 10\log_{10}(1.47\times 10^{8}) \approx 81.7\,\si{\decibel}.
\]
The background sits about four decibels below the machine. The subtraction
is reliable only while the two terms are within about ten decibels of each
other; when the background is far below the source the difference of two
nearly equal large numbers becomes ill-conditioned, which is the
measurement reason a standard requires the source to exceed the background
by at least a few decibels before a corrected level is reported at all.

\subsection{Beat frequency and the tuning of two sources}

Two pure tones close in frequency do not stay separate; they fuse into one
tone that swells and fades, and the rate of the swell is the tool a piano
tuner and a turbine balancer both use. Two loudspeakers emit
$f_1 = 440\,\si{\hertz}$ and $f_2 = 443\,\si{\hertz}$. Find the perceived
tone, the beat rate, and the time between loudness maxima.

\textbf{Superposition.} Adding two equal-amplitude cosines and applying
the sum-to-product identity,
\[
\cos(2\pi f_1 t) + \cos(2\pi f_2 t)
= 2\cos\!\big(2\pi \tfrac{f_1-f_2}{2}t\big)\,
   \cos\!\big(2\pi \tfrac{f_1+f_2}{2}t\big),
\]
a fast carrier at the mean frequency $\tfrac12(f_1+f_2) = 441.5\,\si{
\hertz}$ modulated by a slow envelope at the difference half-frequency
$\tfrac12|f_1-f_2| = 1.5\,\si{\hertz}$.

\textbf{Beat rate.} The ear hears loudness, which follows the squared
envelope, and the envelope magnitude $|\cos(2\pi\cdot 1.5\,t)|$ peaks
twice per envelope cycle, so the loudness pulses at
\[
f_{\text{beat}} = |f_1 - f_2| = 3\,\si{\hertz},
\]
three swells per second, one every $0.33\,\si{\second}$. A piano tuner
slows the beats to zero to bring a string to a reference; a machinery
analyst hears beats between two shafts running at nearly the same speed and
reads the small speed difference straight off the beat rate, no tachometer
required.

\subsection{Critical frequency and the coincidence dip of a panel}

The mass law of the noise-control case study is optimistic at one
frequency, where a bending wave in the panel matches the trace of the
incident sound wave and the panel goes nearly transparent. Find the
coincidence (critical) frequency of a $6\,\si{\milli\meter}$ glass pane and
explain the dip it puts in the panel's transmission loss.

\textbf{The coincidence condition.} A panel carries bending waves whose
speed rises with frequency, $c_B = \sqrt{\omega}\,(B/m'')^{1/4}$ with $B$
the bending stiffness per unit width and $m''$ the surface mass. An
airborne wave arriving at grazing incidence has a trace speed along the
panel equal to the sound speed $c$. When the bending-wave speed equals the
airborne trace speed the panel resonates with the field and transmits
freely; the lowest frequency at which this is possible (at grazing
incidence) is the critical frequency
\[
f_c = \frac{c^2}{2\pi}\sqrt{\frac{m''}{B}}
    = \frac{c^2}{2\pi}\sqrt{\frac{12\rho(1-\nu^2)}{E h^2}},
\]
using $B = E h^3/[12(1-\nu^2)]$ and $m'' = \rho h$ for a plate of
thickness $h$.

\textbf{Glass numbers.} For glass take $E = 70\,\si{\giga\pascal}$,
$\rho = 2500\,\si{\kilogram\per\meter\cubed}$, $\nu = 0.22$,
$h = 6\,\si{\milli\meter}$. The grouped factor is
\[
\sqrt{\frac{12(2500)(1-0.22^2)}{70\times 10^{9}\,(0.006)^2}}
= \sqrt{\frac{12(2500)(0.952)}{70\times 10^{9}(3.6\times 10^{-5})}}
= \sqrt{\frac{2.86\times 10^{4}}{2.52\times 10^{6}}}
= \sqrt{1.13\times 10^{-2}} = 0.107,
\]
so
\[
f_c = \frac{343^2}{2\pi}(0.107)
    = \frac{1.177\times 10^{5}}{6.283}(0.107)
    \approx 2.0\times 10^{3}\,\si{\hertz}.
\]
The pane goes soft near $2\,\si{\kilo\hertz}$, where its measured
transmission loss dips well below the mass-law line, a notch that lands in
the most speech-relevant octave and is why single glazing isolates speech
poorly. The cure is to break the coincidence: a thicker pane lowers $f_c$
out of the band, a laminated pane adds damping that fills the dip, and a
double-glazed unit pairs panes of different thickness so their critical
frequencies do not coincide, the standard reason acoustic glazing uses
unequal leaves.

\subsection{Absorption coefficient from an impedance-tube standing wave}

The reverberation-room example measured a random-incidence absorption on a
large sample. The impedance tube measures the normal-incidence coefficient
on a small one, fast and with little material, by reading the standing-wave
pattern a tone sets up in front of the sample. A rigid tube of inner
diameter $50\,\si{\milli\meter}$ is driven by a loudspeaker at one end and
terminated by a $30\,\si{\milli\meter}$ slab of porous foam backed by a rigid
plug at the other. A probe microphone traversed along the tube at
$500\,\si{\hertz}$ reads a pressure maximum of $2.4\,\si{\pascal}$ and, a
short distance away, a minimum of $0.8\,\si{\pascal}$. Find the sample's
normal-incidence absorption coefficient and the phase of its reflection.

\textbf{The standing-wave ratio.} The incident wave of amplitude $p_i$
running toward the sample and the reflected wave of amplitude $p_r=|R|p_i$
running back superpose to a standing pattern whose envelope ranges between a
maximum $p_{\max}=p_i(1+|R|)$ where the two add and a minimum
$p_{\min}=p_i(1-|R|)$ where they subtract. The measured ratio is
\[
s = \frac{p_{\max}}{p_{\min}} = \frac{2.4}{0.8} = 3.0 ,
\]
the \engterm{standing-wave ratio}, from which the reflection magnitude
follows by inverting $s=(1+|R|)/(1-|R|)$:
\[
|R| = \frac{s-1}{s+1} = \frac{3-1}{3+1} = 0.50 .
\]

\textbf{Absorption coefficient.} The fraction of incident intensity not
returned is the normal-incidence absorption coefficient,
\[
\alpha = 1 - |R|^2 = 1 - 0.25 = 0.75 ,
\]
the marked point of figure~\ref{fig:vol03ch07:impedance-tube}. Three
quarters of the incident energy is absorbed at $500\,\si{\hertz}$.

\textbf{Reading the phase.} The position of the first pressure minimum
relative to the sample face fixes the phase angle of $R$, and through it the
sample's surface impedance $Z_s = Z_0(1+R)/(1-R)$. A minimum exactly at the
face means a real, negative $R$ (a soft, mass-like surface); a minimum a
quarter wavelength out means a real, positive $R$ (a stiff surface). The two
numbers $|R|$ and its phase are the full normal-incidence description, and
the tube delivers both from a single traverse. The coefficient measured here
sits below the random-incidence Sabine value the same foam would show in a
reverberation chamber, because oblique rays strike a porous layer more
effectively than the normal ray; the two standardised methods bracket the
material's behaviour rather than agreeing.

\input{volumes/03-force/ch07-acoustics/figures/fig-impedance-tube.tex}

\subsection{The Newton-Laplace correction to the speed of sound}

The opening derivation fixed the acoustic stiffness as the adiabatic bulk
modulus, not the isothermal one, and the numerical gap between the two is
the century-long error the chapter's history alludes to. Compute both
predictions for air at $0\,{}^{\circ}\mathrm{C}$ and show that only the
adiabatic value matches measurement.

\textbf{Newton's isothermal estimate.} Newton assumed the compressions and
rarefactions of a sound wave stay at constant temperature, so the relevant
stiffness is the isothermal bulk modulus, which for an ideal gas equals the
static pressure, $K_T = p_0$. That gives
\[
c_{\text{iso}} = \sqrt{\frac{p_0}{\rho_0}}
             = \sqrt{\frac{1.013\times 10^5}{1.293}}
             \approx 280\,\si{\meter\per\second} ,
\]
using the density of air at $0\,{}^{\circ}\mathrm{C}$. The measured value is
about $331\,\si{\meter\per\second}$, so Newton's figure is low by roughly
fifteen percent, a discrepancy he could not close and papered over with an
ad hoc correction.

\textbf{Laplace's adiabatic correction.} Laplace recognised that the
compressions are too fast for heat to flow, so the process is adiabatic and
the stiffness is the adiabatic bulk modulus $K_s = \gamma p_0$, larger than
the isothermal modulus by the specific-heat ratio $\gamma = 1.4$ for a
diatomic gas. The speed rises by $\sqrt\gamma$:
\[
c_{\text{adia}} = \sqrt{\frac{\gamma p_0}{\rho_0}}
              = \sqrt{1.4}\times 280
              \approx 331\,\si{\meter\per\second} ,
\]
matching measurement. The factor $\sqrt{1.4}=1.18$ is exactly the fifteen
percent Newton was missing.

\textbf{The lesson in the thermodynamic path.} The choice between two bulk
moduli is a choice of thermodynamic path, and the sound wave selects the
adiabatic one because its period is short compared with the time for heat to
diffuse across a wavelength. The same reasoning fixes the stiffness in every
later formula: the room-mode speed, the impedance $\rho_0 c$, and the
transmission-loss frequencies all carry the adiabatic $c$. A gas whose
relaxation times approach the acoustic period (a molecular gas near a
vibrational resonance) shows the speed drift between the two limits with
frequency, the dispersion that the molecular-relaxation absorption of
section~7.4 accompanies.

\subsection{Steady-state room level and the critical distance}

The room examples so far tracked how sound decays after a source stops. A
running source instead fills the room to a steady level, and predicting that
level is the daily calculation of machine-in-a-room noise. A machine of
sound power level $L_W = 95\,\si{\decibel}$ sits on the floor of a workshop
of volume $V = 400\,\si{\meter\cubed}$ and total surface
$S = 340\,\si{\meter\squared}$ with mean absorption $\bar\alpha = 0.12$. Find
the steady sound pressure level at an operator $1.5\,\si{\meter}$ away and at
a supervisor $8\,\si{\meter}$ away, and locate the critical distance.

\textbf{The room constant.} The reverberant field is set by the
\engterm{room constant}
\[
R = \frac{S\bar\alpha}{1-\bar\alpha}
  = \frac{340\times 0.12}{1-0.12}
  = \frac{40.8}{0.88} \approx 46\,\si{\meter\squared},
\]
the effective area of perfect absorption that would produce the same
reverberant level. A live room has a small $R$ and a loud reverberant field;
a dead room has a large $R$.

\textbf{The two-field level formula.} At distance $r$ from a source of
directivity factor $Q$ (here $Q=2$ for a source on a hard floor radiating
into a half-space), the steady level combines the direct and reverberant
fields on an energy basis,
\[
L_p = L_W + 10\log_{10}\!\left(\frac{Q}{4\pi r^2} + \frac{4}{R}\right),
\]
the two terms of figure~\ref{fig:vol03ch07:room-field}. At the operator
station,
\[
L_p(1.5) = 95 + 10\log_{10}\!\left(\frac{2}{4\pi(1.5)^2} + \frac{4}{46}\right)
        = 95 + 10\log_{10}(0.0707 + 0.087)
        \approx 95 - 8.0 = 87\,\si{\decibel},
\]
and at the supervisor,
\[
L_p(8) = 95 + 10\log_{10}\!\left(\frac{2}{4\pi(8)^2} + \frac{4}{46}\right)
      = 95 + 10\log_{10}(0.00249 + 0.087)
      \approx 95 - 10.5 = 84.5\,\si{\decibel}.
\]
The supervisor, more than five times farther, is only $2.5\,\si{\decibel}$
quieter, because both stations sit in the reverberant field where distance
buys almost nothing.

\textbf{The critical distance.} The direct and reverberant terms are equal
at the critical distance
\[
r_c = \sqrt{\frac{Q R}{16\pi}}
    = \sqrt{\frac{2\times 46}{16\pi}}
    \approx 1.35\,\si{\meter},
\]
so the operator sits barely beyond $r_c$ and the supervisor is deep in the
reverberant field. The design reading is direct: moving the operator back
would not help, because the level is fixed by the room, and the only lever
on the reverberant field is more absorption. Doubling $R$ from $46$ to
$92\,\si{\meter\squared}$ lowers the reverberant plateau by
$10\log_{10}2 = 3\,\si{\decibel}$ and pushes $r_c$ out to
$1.9\,\si{\meter}$, which is why a workshop noise cure treats the ceiling
before it relocates the worker.

\input{volumes/03-force/ch07-acoustics/figures/fig-room-field.tex}

\subsection{Daily noise dose from a variable-level shift}

The enclosure case study scored a single steady level against the exposure
limit. A real shift mixes tasks at different levels, and the correct
accounting sums partial doses rather than averaging levels. A machinist
spends a shift across three tasks: $6\,\si{\hour}$ at a bench reading
$82\,\si{\decibel}$(A), $1.5\,\si{\hour}$ at a grinder reading
$92\,\si{\decibel}$(A), and $0.5\,\si{\hour}$ at a chipping station reading
$100\,\si{\decibel}$(A). Find the daily dose and the equivalent eight-hour
level, and identify the task that controls the day.

\textbf{Permitted time per task.} Under the equal-energy ($3\,\si{\decibel}$)
rule with the $85\,\si{\decibel}$(A) eight-hour criterion, the permitted
time at level $L$ is $T_i = 8/2^{(L-85)/3}\,\si{\hour}$, the blue curve of
figure~\ref{fig:vol03ch07:noise-dose}. The three tasks give
\[
T_{82} = 8\cdot 2^{3/3} = 16\,\si{\hour},\quad
T_{92} = 8\cdot 2^{-7/3} = 1.59\,\si{\hour},\quad
T_{100} = 8\cdot 2^{-15/3} = 0.25\,\si{\hour}.
\]

\textbf{Summing partial doses.} The dose is the sum of the fractions of each
permitted time actually used,
\[
D = \sum_i \frac{t_i}{T_i}
  = \frac{6}{16} + \frac{1.5}{1.59} + \frac{0.5}{0.25}
  = 0.375 + 0.943 + 2.0 = 3.32 ,
\]
that is $332\,\%$ of the permissible daily dose, more than three times the
limit. The half-hour of chipping alone contributes a dose of $2.0$, twice
the whole day's allowance, while the six hours at the bench contribute a
third of it.

\textbf{Equivalent eight-hour level.} The dose converts to an equivalent
continuous level by $L_{\text{EX,8h}} = 85 + 10\log_{10}D = 85 +
10\log_{10}(3.32) \approx 90.2\,\si{\decibel}$(A). A control strategy reads
straight off the partial doses: halving the chipping time, or a
$5\,\si{\decibel}$ enclosure on the chipping station, removes more dose than
silencing the bench entirely. The transportable habit is that dose is a
time integral dominated by its loudest, briefest term, so the day is
controlled by the task a level-only survey would dismiss as too short to
matter.

\input{volumes/03-force/ch07-acoustics/figures/fig-noise-dose.tex}

\subsection{Ground-reflection interference over hard ground}

The free-field and line-source formulas ignored the ground. Outdoors the
ground returns a second path, and its interference with the direct path cuts
deep frequency-dependent notches into the received spectrum, the effect that
makes an outdoor measurement depend on microphone height. A source sits
$0.5\,\si{\meter}$ above hard ground and a receiver $1.2\,\si{\meter}$ above
it, $20\,\si{\meter}$ away horizontally. Find the path-length difference
between the direct and ground-reflected rays and the lowest frequency at
which the two cancel.

\textbf{The two paths.} The direct path is the straight line between source
and receiver,
\[
d_{\text{dir}} = \sqrt{20^2 + (1.2-0.5)^2} = \sqrt{400 + 0.49}
             \approx 20.012\,\si{\meter}.
\]
The reflected path is found by the image source, the source mirrored below
the ground plane at $-0.5\,\si{\meter}$, so
\[
d_{\text{ref}} = \sqrt{20^2 + (1.2+0.5)^2} = \sqrt{400 + 2.89}
             \approx 20.072\,\si{\meter}.
\]
The path difference is
\[
\delta = d_{\text{ref}} - d_{\text{dir}} \approx 0.060\,\si{\meter}.
\]

\textbf{The first cancellation.} A hard ground reflects with no phase change,
so the two rays arrive in antiphase, and cancel, when the path difference is
half a wavelength, $\delta = \lambda/2$, that is at
\[
f_{\text{null}} = \frac{c}{2\delta} = \frac{343}{2(0.060)}
              \approx 2.86\,\si{\kilo\hertz}.
\]
Below this the two paths reinforce toward the $+6\,\si{\decibel}$ in-phase
sum; above it the level combs through further nulls at odd multiples of
$f_{\text{null}}$, the pattern of figure~\ref{fig:vol03ch07:ground-interference}
(drawn for a larger path difference so the comb is visible on the plotted
band).

\textbf{Why measurement standards fix the heights.} The null frequency
depends only on the path difference, which the source and receiver heights
set, so raising the microphone slides the whole comb. A measurement made at
one height reports a spectrum riddled with nulls that a different height
would place elsewhere, which is why standardised outdoor and facade
measurements fix the microphone height and why a single-point outdoor
reading of a tonal source is unreliable near a null. Over soft ground the
reflection is weaker and carries a frequency-dependent phase shift, which
fills the sharp nulls and replaces them with a broad low-frequency excess
attenuation, the ground effect that community-noise models tabulate by
ground type.

\input{volumes/03-force/ch07-acoustics/figures/fig-ground-interference.tex}

\subsection{Pulse compression: resolution without losing range}

The sonar and NDT examples met the pulse-length trade head-on: a short pulse
resolves finely but carries little energy. The chirp escapes the trade, and
quantifying the escape shows why every modern ranging system transmits a
coded pulse. A sonar transmits a linear-frequency-sweep (chirp) pulse of
duration $\tau = 10\,\si{\milli\second}$ that sweeps across a bandwidth
$B = 20\,\si{\kilo\hertz}$ in water. Find the range resolution before and
after matched-filter compression and the processing gain.

\textbf{Uncompressed resolution.} A plain pulse of duration $\tau$ occupies a
spatial length $c\tau$, and two reflectors are resolved only if separated by
half that, so
\[
\Delta r_{\text{plain}} = \frac{c\tau}{2}
                       = \frac{1480\times 0.010}{2}
                       = 7.4\,\si{\meter}.
\]
A ten-millisecond pulse cannot tell apart two targets closer than seven
metres, useless for imaging.

\textbf{Compressed resolution.} Matched filtering a chirp collapses it to an
effective duration $\tau_c \approx 1/B$, so the compressed range resolution
is set by the bandwidth, not the pulse length,
\[
\Delta r_{\text{comp}} = \frac{c}{2B}
                      = \frac{1480}{2(20\times 10^3)}
                      \approx 0.037\,\si{\meter},
\]
about four centimetres, a factor of two hundred finer than the plain pulse
while radiating the same energy over the same ten milliseconds.

\textbf{The time-bandwidth product.} The compression ratio is the
\engterm{time-bandwidth product}
\[
\tau B = (10\times 10^{-3})(20\times 10^3) = 200 ,
\]
which is both the resolution improvement and the signal-to-noise processing
gain, $10\log_{10}(200)\approx 23\,\si{\decibel}$, that the matched filter
recovers by summing the long pulse coherently. The design escape is now
explicit: range comes from the total pulse energy $\propto\tau$, resolution
comes from the bandwidth $B$, and the chirp decouples them so a system
chooses each independently. Radar, medical ultrasound, and seismic
vibroseis all run on the same time-bandwidth trade, and the only cost is a
matched-filter sidelobe floor that windowing trims.

\subsection{Classical absorption length of an ultrasonic beam in air}

The attenuation section gave the classical coefficient its $\omega^2$ law
and a working constant; here we turn that constant into the range a beam
survives, the number that decides whether an airborne ultrasonic sensor can
reach its target. Find the distance over which the pressure amplitude of an
airborne ultrasonic beam falls to one tenth of its launched value, at
$40\,\si{\kilo\hertz}$ (a proximity-sensor frequency) and at
$1\,\si{\mega\hertz}$, from the classical absorption alone.

\textbf{From coefficient to length.} The classical coefficient for air is
$\alpha_{\mathrm{cl}}\approx 1.4\times 10^{-7}\,f^2\,\si{\decibel\per\kilo
\meter}$ with $f$ in hertz. A fall to one tenth of the amplitude is a drop
of $20\,\si{\decibel}$, so the ten-percent range is $r_{10} =
20/\alpha_{\mathrm{cl}}$ in kilometres.

\textbf{The two frequencies.} At $40\,\si{\kilo\hertz}$,
\[
\alpha_{\mathrm{cl}} = 1.4\times 10^{-7}(4\times 10^{4})^2
                     = 1.4\times 10^{-7}(1.6\times 10^{9})
                     \approx 2.2\times 10^{2}\,\si{\decibel\per\kilo\meter},
\]
so $r_{10} = 20/(2.2\times 10^{2}) \approx 9\times 10^{-2}\,\si{\kilo\meter}
\approx 90\,\si{\meter}$. Classical absorption alone is gentle here: a
$40\,\si{\kilo\hertz}$ proximity beam would keep a tenth of its amplitude
out to tens of metres if the thermoviscous mechanism were the only loss.
Real range-finders are held to a few metres instead by molecular
relaxation, which adds a humidity-dependent excess far above this classical
floor, and by geometric spreading; the classical figure sets the optimistic
scale, not the achieved range. At $1\,\si{\mega\hertz}$,
\[
\alpha_{\mathrm{cl}} = 1.4\times 10^{-7}(10^{6})^2
                     = 1.4\times 10^{5}\,\si{\decibel\per\kilo\meter}
                     = 1.4\times 10^{2}\,\si{\decibel\per\meter},
\]
so the amplitude falls by $20\,\si{\decibel}$ in
$r_{10} = 20/(1.4\times 10^{2}) \approx 0.14\,\si{\meter}$,
about fourteen centimetres. Airborne megahertz ultrasound barely clears a
hand's width, the reason every megahertz application couples into a liquid or a
solid, where the impedance is high and the absorption per wavelength is far
lower.

\textbf{What the example fixes in habit.} The $f^2$ absorption law is not a
mild correction but the hard wall that partitions ultrasonics by medium:
tens of kilohertz for airborne ranging over metres, megahertz only in
water or tissue over centimetres. Doubling the frequency quarters the
range at fixed launched level, so an ultrasonic system's operating
frequency is chosen against the absorption curve of
figure~\ref{fig:vol03ch07:classical-absorption} before any transducer is
selected.

\subsection{Rayleigh distance and the near field of a piston}

The spherical-spreading law assumed a compact source, and the sound-power
survey warned that a microphone in the near field reads a pressure that does
not represent radiated power. A flat radiator has a near field far longer
than a wavelength, and knowing where it ends decides where a measurement or
a focus can be trusted. Find the extent of the near field of a
$50\,\si{\milli\meter}$-diameter ultrasonic transducer radiating at
$2\,\si{\mega\hertz}$ into water, and the beam's far-field divergence
angle.

\textbf{The Rayleigh distance.} A circular piston of radius $a$ radiating a
wavelength $\lambda$ has a near field of on-axis maxima and minima that ends
at the last axial maximum, the \engterm{Rayleigh distance}
\[
N = \frac{a^2}{\lambda} = \frac{D^2}{4\lambda},
\]
with $D=2a$ the diameter. In water at $2\,\si{\mega\hertz}$ the wavelength is
$\lambda = c/f = 1480/(2\times 10^{6}) = 7.4\times 10^{-4}\,\si{\meter}$, so
\[
N = \frac{(0.025)^2}{7.4\times 10^{-4}}
  = \frac{6.25\times 10^{-4}}{7.4\times 10^{-4}}
  \approx 0.84\,\si{\meter}.
\]
The near field extends most of a metre. Inside it the on-axis pressure
oscillates between zero and twice the far-field value, so a reflector or a
hydrophone placed there sees a reading that does not obey the inverse
distance law, and a focus set inside $N$ is limited by the piston's natural
last maximum rather than by geometry.

\textbf{The far-field divergence.} Beyond the Rayleigh distance the beam
spreads into a cone whose half-angle to the first null follows the piston
directivity,
\[
\sin\theta_{\text{null}} = 1.22\,\frac{\lambda}{D}
                        = 1.22\,\frac{7.4\times 10^{-4}}{0.050}
                        \approx 0.0181,
\]
so $\theta_{\text{null}}\approx 1.0^{\circ}$, a tight beam. A larger
aperture or a shorter wavelength narrows the cone and lengthens the near
field together, the same $D/\lambda$ that set the directivity of
figure~\ref{fig:vol03ch07:directivity}.

\textbf{What the example fixes in habit.} A flat radiator has two ranges,
not one: a near field of length $a^2/\lambda$ where the on-axis pressure is
non-monotonic and untrustworthy, and a far field where the inverse-distance
law and a fixed divergence angle apply. Ultrasound focusing, transducer
qualification, and the far-field condition for a sound-power survey all read
off the same Rayleigh distance, so it is computed before a probe is aimed.

\subsection{Steering a line array and the grating-lobe limit}

The directivity example fixed the beam of a single piston; a line of small
sources beats it by steering the beam electronically, the working principle
of a phased sonar, a medical-imaging array, and a steerable public-address
column. Find the element spacing that lets a line array steer to
$45^{\circ}$ without a spurious second beam, at $3\,\si{\kilo\hertz}$ in
air, and the main-lobe width of a sixteen-element array at that spacing.

\textbf{The steering phase.} Delaying each element by a progressive time
tilts the wavefront. A phase increment $\psi = kd\sin\theta_0$ between
adjacent elements steers the main beam to angle $\theta_0$, where $d$ is the
element spacing and $k=2\pi/\lambda$. The array factor repeats whenever the
phase across one spacing advances by a full $2\pi$, which plants a
\engterm{grating lobe}, a full-strength copy of the main beam, at an angle
$\theta_g$ satisfying $\sin\theta_g = \sin\theta_0 - \lambda/d$.

\textbf{The spacing that suppresses the grating lobe.} To keep the grating
lobe out of the visible range $|\sin\theta_g|\le 1$ for a beam steered to
$\theta_0$, the spacing must satisfy
\[
\frac{d}{\lambda} \le \frac{1}{1+|\sin\theta_0|}.
\]
At $3\,\si{\kilo\hertz}$ the wavelength is $\lambda = 343/3000 =
0.114\,\si{\meter}$, and steering to $\theta_0 = 45^{\circ}$
($\sin\theta_0 = 0.707$) requires
\[
d \le \frac{0.114}{1.707} \approx 0.067\,\si{\meter},
\]
so elements no more than about $67\,\si{\milli\meter}$ apart. A wider
spacing is cheaper and gives a narrower beam but lets a second beam appear
in an unwanted direction, the classic aliasing trap of a sampled aperture.

\textbf{The main-lobe width.} A uniform array of $N$ elements at spacing $d$
has a main-lobe half-power width, for a broadside beam, of about
\[
\Delta\theta \approx \frac{0.886\,\lambda}{Nd}\ \text{radians}
             = \frac{0.886(0.114)}{16(0.067)}
             \approx 0.094\,\text{rad} \approx 5.4^{\circ}.
\]
Steering off broadside broadens the beam by $1/\cos\theta_0$, so at
$45^{\circ}$ the width grows to about $5.4/\cos 45^{\circ}\approx
7.6^{\circ}$. A longer array (more elements at the same spacing) narrows the
beam in proportion to $Nd$, the array's total aperture.

\textbf{What the example fixes in habit.} An array trades a fixed aperture
for a steerable beam, and the spacing is bounded above by the grating-lobe
condition, not chosen freely: half a wavelength is the safe default that
steers anywhere without aliasing. Beamwidth is set by the total aperture
$Nd$ and broadens as the beam steers off broadside, the two facts that size
every phased array in the chapter's ranging and imaging uses.

\subsection{Loudspeaker sensitivity and electroacoustic efficiency}

The radiator worked example converted a cone motion into an on-axis level;
a buyer reads a driver instead by its sensitivity and its efficiency, and
the two numbers expose how little of the amplifier's power becomes sound.
A loudspeaker is rated at a sensitivity of $88\,\si{\decibel}$ SPL at
$1\,\si{\meter}$ for $1\,\si{\watt}$ of electrical input. Find its acoustic
output power and its electroacoustic efficiency, and the amplifier power
needed to reach $104\,\si{\decibel}$ at the listener $3\,\si{\meter}$ away.

\textbf{From sensitivity to radiated power.} The driver on a baffle
radiates into a half-space, so at $1\,\si{\meter}$ its acoustic power $W$
gives an intensity $I = W/(2\pi r^2)$ and a level $L_p = L_W - 10\log_{10}
(2\pi r^2)$. Inverting from the rated $88\,\si{\decibel}$ at $1\,\si{\meter}$,
\[
L_W = L_p + 10\log_{10}(2\pi r^2) = 88 + 10\log_{10}(6.28) \approx
      88 + 8.0 = 96\,\si{\decibel},
\]
and with the power reference $W_0 = 10^{-12}\,\si{\watt}$,
\[
W = W_0\,10^{L_W/10} = 10^{-12}\times 10^{9.6}
  \approx 4.0\times 10^{-3}\,\si{\watt}.
\]

\textbf{The efficiency.} Four milliwatts of sound came out of one watt of
electrical input, so the electroacoustic efficiency is
\[
\eta = \frac{W_{\text{ac}}}{W_{\text{elec}}}
     = \frac{4.0\times 10^{-3}}{1}
     \approx 0.4\,\%,
\]
a typical figure for a direct-radiator cone driver. Better than
ninety-nine percent of the amplifier's power heats the voice coil rather
than radiating, which is why high-output systems favour horn loading (a few
percent efficient) over raw amplifier power.

\textbf{The power to reach the target level.} At $3\,\si{\meter}$ the
half-space level for input power $P$ is $L_p = 88 + 10\log_{10}P -
20\log_{10}(3/1)$, since doubling distance in the direct field costs
$6\,\si{\decibel}$. Setting $L_p = 104\,\si{\decibel}$,
\[
10\log_{10}P = 104 - 88 + 20\log_{10}3 = 16 + 9.5 = 25.5,
\]
so $P = 10^{2.55}\approx 355\,\si{\watt}$. Reaching a loud but not extreme
level at a normal listening distance already asks for hundreds of watts,
the direct consequence of the sub-percent efficiency, and the reason a
three-decibel sensitivity gain (which halves the required amplifier power)
is worth more than a large increase in amplifier rating.

\textbf{What the example fixes in habit.} A loudspeaker's sensitivity and
its efficiency are two readings of the same fact, that a cone driver turns
under a percent of its electrical input into sound, so headroom is bought
by sensitivity, by half-space or horn loading, and by proximity, far more
cheaply than by amplifier power. The $6\,\si{\decibel}$-per-doubling direct
field and the half-space power relation of the sound-power survey are the
two levers that connect the rating to the room.

\subsection{Shock-formation distance of a high-amplitude wave}

The nonlinear mastery box named the steepening of a loud wave into a shock;
here we compute where the shock forms, the number that tells a siren or a
sonic-boom designer whether the wave has already broken by the time it
reaches the listener. Find the shock-formation distance for a
$1\,\si{\kilo\hertz}$ plane tone in air at a source level of
$155\,\si{\decibel}$, and compare it to the same tone at
$120\,\si{\decibel}$.

\textbf{The steepening mechanism.} The local wave speed is $c + \beta u$,
with $u$ the particle velocity and $\beta = (\gamma+1)/2 = 1.2$ for air the
coefficient of nonlinearity, so the crests outrun the troughs and the
waveform steepens. The distance at which the leading edge becomes vertical,
the shock-formation distance, is
\[
x_{\text{shock}} = \frac{c^2}{\beta\,\omega\,\hat u}
                 = \frac{\rho_0 c^3}{\beta\,\omega\,\hat p},
\]
using $\hat u = \hat p/(\rho_0 c)$.

\textbf{The two levels.} At $155\,\si{\decibel}$ the peak pressure is, from
$L = 20\log_{10}(p_{\text{rms}}/p_{\text{ref}})$ with $p_{\text{ref}} =
20\,\si{\micro\pascal}$, an rms of $p_{\text{rms}} = 2\times 10^{-5}\times
10^{155/20}\approx 1125\,\si{\pascal}$, so $\hat p = \sqrt2(1125)\approx
1590\,\si{\pascal}$. With $\rho_0 c^3 = 1.21(343)^3\approx 4.88\times
10^{7}$ and $\omega = 2\pi(1000) = 6283\,\si{\radian\per\second}$,
\[
x_{\text{shock}} = \frac{4.88\times 10^{7}}{1.2(6283)(1590)}
                 \approx \frac{4.88\times 10^{7}}{1.20\times 10^{7}}
                 \approx 4\,\si{\meter}.
\]
The wave breaks into a sawtooth within a few metres of a
$155\,\si{\decibel}$ source. At $120\,\si{\decibel}$ the peak pressure is
smaller by a factor $10^{35/20}\approx 56$, so $\hat p\approx
28\,\si{\pascal}$ and the shock distance grows to
$x_{\text{shock}}\approx 4\times 56 \approx 230\,\si{\meter}$, far enough
that ordinary loud sound stays linear over any room and most of the
outdoors.

\textbf{What the example fixes in habit.} Shock formation is a threshold in
amplitude, not a gradual effect: below roughly $150\,\si{\decibel}$ the
shock distance runs to hundreds of metres and the linear theory of the whole
chapter holds, while above it the distance collapses into metres and the
wave arrives as a shock with the excess absorption and harmonic content the
nonlinear box describes. Sirens, blast waves, and sonic booms live on the
short-distance side, which is why they are treated with nonlinear acoustics
and ordinary noise control is not.

\subsection{Acoustic radiation force and a standing-wave levitator}

Every example so far treated sound as a carrier of information or energy; a
standing wave also exerts a steady force, strong enough to suspend small
objects against gravity, the working principle of the acoustic levitator
used to hold droplets for containerless processing. Find the sound pressure
level a standing wave needs to levitate a small water droplet in air, and
the drop size the method can hold.

\textbf{The radiation force.} A standing wave of pressure amplitude $\hat p$
holds a small dense particle near a pressure node with a restoring force
whose peak, for a particle of radius $R$ small compared to the wavelength,
scales as
\[
F_{\text{rad}} \sim \frac{5}{6}\,\pi R^3 k\,\frac{\hat p^2}{2\rho_0 c^2},
\]
the acoustic-contrast factor for a dense particle in air being of order
unity. Levitation requires this peak force to exceed the particle weight
$mg = \tfrac43\pi R^3\rho_p g$.

\textbf{The required level.} Setting $F_{\text{rad}} = mg$ and cancelling the
common $\pi R^3$ leaves a condition independent of drop size,
\[
\frac{5}{6}\,k\,\frac{\hat p^2}{2\rho_0 c^2}
   = \frac{4}{3}\rho_p g,
\qquad
\hat p^2 = \frac{16\,\rho_p g\,\rho_0 c^2}{5\,k}.
\]
Take a levitator at $f = 20\,\si{\kilo\hertz}$, so $k = 2\pi f/c =
2\pi(20000)/343 \approx 366\,\si{\per\meter}$, with $\rho_p =
1000\,\si{\kilogram\per\meter\cubed}$ for water and $\rho_0 c^2 =
\gamma p_0 = 1.42\times 10^{5}\,\si{\pascal}$. Then
\[
\hat p^2 = \frac{16(1000)(9.81)(1.42\times 10^{5})}{5(366)}
         \approx 1.22\times 10^{7}\,\si{\pascal\squared},
\]
so $\hat p\approx 3.5\times 10^{3}\,\si{\pascal}$, an rms of about
$2.5\times 10^{3}\,\si{\pascal}$, which is
\[
L = 20\log_{10}\frac{2.5\times 10^{3}}{2\times 10^{-5}}
  \approx 162\,\si{\decibel}.
\]
A standing wave near $160\,\si{\decibel}$ levitates the droplet, a level
reached inside a tuned resonator between an ultrasonic horn and a reflector
even though no such level is safe in the open.

\textbf{The size limit.} The small-particle scaling held only while
$R\ll\lambda$; at $20\,\si{\kilo\hertz}$ the wavelength is
$\lambda = 343/20000 = 17\,\si{\milli\meter}$, so the method holds drops up
to a few millimetres, above which the particle is no longer small compared
to the wavelength and the simple force law fails. The levitated drop sits
slightly below the pressure node, where the radiation force balances
gravity.

\textbf{What the example fixes in habit.} Radiation force is a second-order,
time-averaged effect that survives when the first-order oscillation averages
to zero, so it scales as the square of the pressure and needs the very high
levels a closed resonator supplies. The same standing-wave force sorts cells
in a microfluidic channel and drives acoustic tweezers, all read from the
node structure of the field rather than from its instantaneous pressure.

\section{Project}\pathtag{core}

\begin{project}[Characterise the impulse response of a household
room]
The reader measures the acoustic response of a household room
(living room, bedroom, hallway) to broadband excitation, extracts
the reverberation time $T_{60}$ as a function of frequency,
identifies the lowest few modal frequencies, and compares to
theoretical predictions from the room's geometry. The deliverable is
a twenty- to thirty-page engineering report.

\textbf{Track}. Physical measurement plus analysis. \textbf{Hazard
class}: low. Use hearing protection if the source is loud. Avoid
explosive noise sources; use clapping or a chirp from a speaker.

\textbf{Apparatus}. A microphone (a smartphone's built-in microphone
with a free audio-recording app is sufficient for first-cut measurements;
a calibrated measurement microphone is preferable). A measurement
location (the room being characterised). A noise source: a single
loud clap, a balloon pop (for impulse response), or a calibrated
speaker playing a frequency sweep or pink noise.

\textbf{Procedure}. The reader records the room's response to (a)
a single broadband impulse (clap or balloon pop), and (b) a slow
sine-frequency sweep from $20\,\si{\hertz}$ to $20\,\si{\kilo\hertz}$
played through a speaker. Recordings should be from at least three
microphone positions in the room; the loudest position is typically
near a corner (where all modes have a pressure maximum).

\textbf{Analysis}. The reader:

\begin{itemize}[leftmargin=*]
  \item Computes the FFT of the swept-sine recording and identifies
    the lowest five to ten room modes.
  \item Plots the energy decay of the impulse response in $1/3$-octave
    bands and extracts $T_{60}$ at the standard band centres
    ($125\,\si{\hertz}$, $250\,\si{\hertz}$, $500\,\si{\hertz}$,
    $1\,\si{\kilo\hertz}$, $2\,\si{\kilo\hertz}$, $4\,\si{\kilo\hertz}$).
  \item Predicts the lowest five room modes from the room's
    measured dimensions using the rectangular-room formula of
    section 7.5; compares to the measured modes.
  \item Predicts $T_{60}$ from the Sabine formula using the room's
    volume and an estimated total absorption; compares to the
    measured $T_{60}$ at $500\,\si{\hertz}$.
\end{itemize}

\textbf{Reflection ($1{,}500$ to $3{,}000$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
  \item Which measured modes match the rectangular-room prediction;
    which differ and why (non-rectangular geometry, soft surfaces,
    flanking transmission).
  \item How the measured $T_{60}$ varies with frequency, and what
    that variation indicates about the dominant absorbers in the
    room (carpet, drapery, upholstered furniture, ceiling material).
  \item Whether the room would benefit (for music listening,
    speech, or quiet conversation) from additional absorption or
    diffusion.
  \item One feature of the measurement (microphone position,
    background noise, recording-device dynamic range) that limited
    the accuracy.
\end{itemize}

\textbf{Deliverable}. The recorded signals, the FFT plots, the
$T_{60}$ analysis, the predicted-versus-measured mode table, and
the reflection.

\textbf{Time}. Two to four hours of measurement, six to ten hours of
signal-processing analysis, four to eight hours of write-up.
\end{project}

\section{Exercises}

The exercises mix calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
The speed of sound in air at $20^{\circ}\,\mathrm{C}$ is approximately
$343\,\si{\meter\per\second}$. Find the wavelength of (a) a $440\,\si
{\hertz}$ tone, (b) a $20\,\si{\kilo\hertz}$ tone, (c) a $20\,\si{
\hertz}$ tone.
\paragraph{Solution sketch.} Apply $\lambda = c/f$.
(a) $\lambda = 343/440 = 0.78\,\si{\meter}$.
(b) $\lambda = 343/20{,}000 = 0.0172\,\si{\meter} = 17.2\,\si{
\milli\meter}$.
(c) $\lambda = 343/20 = 17.2\,\si{\meter}$. The audible band spans
three orders of magnitude in wavelength, from the width of a desk to
the length of a house.
\end{exercise}

\begin{exercise}
A sound source produces a pressure of $1\,\si{\pascal}$ rms at $1\,\si
{\meter}$ in air. Find the sound pressure level in dB.
\paragraph{Solution sketch.} With $p_{\text{ref}} = 20\,\si{
\micro\pascal} = 2\times 10^{-5}\,\si{\pascal}$,
\[
\text{SPL} = 20\log_{10}\!\frac{1}{2\times 10^{-5}}
= 20\log_{10}(5\times 10^{4}) \approx 20(4.70) = 94\,\si{\decibel}.
\]
The convenient anchor to remember: $1\,\si{\pascal}$ rms is
$94\,\si{\decibel}$ SPL, the level many microphone calibrators emit.
\end{exercise}

\begin{exercise}
Compute the pressure-amplitude reflection coefficient and intensity
transmission coefficient for a sound wave normally incident from
air into water.
\paragraph{Solution sketch.} With $Z_1 = Z_{\text{air}} = 415$ and
$Z_2 = Z_{\text{water}} = 1.5\times 10^{6}\,\text{rayl}$,
\[
R = \frac{Z_2 - Z_1}{Z_2 + Z_1}
= \frac{1.5\times 10^{6} - 415}{1.5\times 10^{6} + 415}
\approx 0.99945.
\]
The intensity transmission is
$T_I = 1 - R^2 = 4Z_1 Z_2/(Z_1+Z_2)^2 \approx 4(415)/(1.5\times
10^{6}) \approx 1.1\times 10^{-3}$, about $0.11\,\%$. Nearly all the
incident energy reflects; the air-water boundary is almost a mirror.
\end{exercise}

\begin{exercise}
A plane wave in air at $20^{\circ}\,\mathrm{C}$ is incident at
$30^{\circ}$ from normal on a layer of warm air at $30^{\circ}\,
\mathrm{C}$. Find the refraction angle.
\paragraph{Solution sketch.} The sound speed scales as $\sqrt{T}$ with
$T$ absolute. $c_1 = 343\,\si{\meter\per\second}$ at $293\,\si{\kelvin}$;
$c_2 = 343\sqrt{303/293} = 349\,\si{\meter\per\second}$ at
$303\,\si{\kelvin}$. Snell's law gives
\[
\sin\theta_2 = \frac{c_2}{c_1}\sin\theta_1
= \frac{349}{343}\sin 30^{\circ} = 1.0175(0.5) = 0.509,
\]
so $\theta_2 = \arcsin(0.509) \approx 30.6^{\circ}$. The wave bends
slightly away from the normal because it speeds up in the warmer
layer, the same refraction that steers sound toward or away from the
ground under temperature gradients.
\end{exercise}

\begin{exercise}
A rectangular room is $5\,\si{\meter} \times 4\,\si{\meter} \times
2.5\,\si{\meter}$. Find the three lowest axial modal frequencies.
\paragraph{Solution sketch.} Axial modes have one nonzero index, so
$f = c/(2L)$ along each axis. Along the $5\,\si{\meter}$ axis,
$f = 343/10 = 34.3\,\si{\hertz}$; along $4\,\si{\meter}$,
$f = 343/8 = 42.9\,\si{\hertz}$; along $2.5\,\si{\meter}$,
$f = 343/5 = 68.6\,\si{\hertz}$. These match the lowest entries of
\texttt{code/room\_modes.py}. The longest dimension gives the lowest
mode.
\end{exercise}

\begin{exercise}
A room of volume $50\,\si{\meter\cubed}$ has a total absorption of
$10\,\si{\meter\squared}$-sabin. Find the reverberation time $T_{60}$
using the Sabine formula.
\paragraph{Solution sketch.}
$T_{60} = 0.161\,V/A = 0.161(50)/10 = 0.805\,\si{\second}$. Adding
absorption shortens reverberation in inverse proportion: doubling $A$
to $20\,\si{\meter\squared}$-sabin would halve $T_{60}$ to
$\approx 0.40\,\si{\second}$.
\end{exercise}

\begin{exercise}
A sonar pulse echoes from a target after $0.6\,\si{\second}$ in
seawater (speed of sound $1{,}500\,\si{\meter\per\second}$). Find the
target's distance.
\paragraph{Solution sketch.} The pulse makes a round trip, so the
target distance is $r = c\,t/2 = 1{,}500(0.6)/2 = 450\,\si{\meter}$.
The factor of two is the recurring trap in pulse-echo ranging: the
measured time is for there and back.
\end{exercise}

\begin{exercise}
A $2.25\,\si{\mega\hertz}$ ultrasonic pulse travels through a
$25\,\si{\milli\meter}$-thick aluminum plate (longitudinal speed
$6{,}300\,\si{\meter\per\second}$). Find the round-trip time and the
spatial resolution of the pulse, given a pulse length of three wave
periods.
\paragraph{Solution sketch.} Round-trip time:
$t = 2L/c = 2(0.025)/6{,}300 = 7.94\,\si{\micro\second}$. The
wavelength is $\lambda = c/f = 6{,}300/(2.25\times 10^{6}) =
2.8\,\si{\milli\meter}$, so the pulse length is
$3\lambda = 8.4\,\si{\milli\meter}$ and the axial resolution (half the
pulse length, to separate two echoes) is $\approx 4.2\,\si{
\milli\meter}$. Shorter pulses, or higher frequencies, resolve finer
features at the cost of penetration depth.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the acoustic wave equation from the linearised conservation
of mass, Euler's equation, and an adiabatic equation of state for an
ideal gas. State the wave speed in terms of the gas's specific
heat ratio and temperature.
\paragraph{Solution sketch.} Write each field as ambient plus a small
perturbation: $\rho = \rho_0 + \rho'$, $p = p_0 + p'$, $\vec{u}$ small.
Linearised mass conservation:
$\partial_t \rho' + \rho_0\nabla\cdot\vec{u} = 0$. Linearised Euler
(inviscid momentum): $\rho_0\,\partial_t\vec{u} = -\nabla p'$. The
isentropic state relation gives $p' = c^2\rho'$ with
$c^2 = (\partial p/\partial\rho)_s$. Take $\partial_t$ of the mass
equation and substitute the divergence of the momentum equation:
$\partial_{tt}\rho' = \nabla^2 p' = c^2\nabla^2\rho'$, and since
$p' = c^2\rho'$ the same equation holds for $p'$, giving
$\partial_{tt}p' = c^2\nabla^2 p'$. For an ideal gas
$c^2 = \gamma p_0/\rho_0 = \gamma R T$, so $c = \sqrt{\gamma R T}$. The
key assumptions are small amplitude (products of perturbations
dropped) and isentropic compression (entropy change negligible on the
acoustic timescale).
\end{exercise}

\begin{exercise}
Derive the longitudinal-wave speed $c_L = \sqrt{(\lambda + 2\mu)/
\rho}$ in an isotropic elastic solid. State the assumption about
the spatial variation of the wave's displacement.
\paragraph{Solution sketch.} Start from the Navier equation of
elastodynamics,
$\rho\,\partial_{tt}\vec{u} = (\lambda+\mu)\nabla(\nabla\cdot\vec{u})
+ \mu\nabla^2\vec{u}$, with $\lambda,\mu$ the Lam\'e constants. Assume a
plane longitudinal wave travelling along $x$ with displacement only in
$x$: $\vec{u} = (u_x(x,t),0,0)$. Then $\nabla\cdot\vec{u} =
\partial_x u_x$ and $\nabla(\nabla\cdot\vec{u}) =
(\partial_{xx}u_x,0,0)$, while $\nabla^2\vec{u} =
(\partial_{xx}u_x,0,0)$. Substituting gives
$\rho\,\partial_{tt}u_x = (\lambda+2\mu)\,\partial_{xx}u_x$, a scalar
wave equation with speed $c_L = \sqrt{(\lambda+2\mu)/\rho}$. The
assumption is that the displacement varies only along the propagation
direction (a one-dimensional plane wave); the corresponding transverse
case, $\vec{u}=(0,u_y(x,t),0)$, drops the dilatation term and yields
$c_T = \sqrt{\mu/\rho}$.
\end{exercise}

\begin{exercise}
Derive the normal-incidence pressure reflection coefficient $R = (Z_2
- Z_1)/(Z_2 + Z_1)$ from the boundary conditions of continuous
pressure and continuous normal velocity at the interface.
\paragraph{Solution sketch.} Write the fields in medium 1 as incident
plus reflected and in medium 2 as transmitted:
$p_1 = p_i e^{i(k_1 x - \omega t)} + p_r e^{i(-k_1 x - \omega t)}$,
$p_2 = p_t e^{i(k_2 x - \omega t)}$. For a plane wave the particle
velocity is $u = \pm p/Z$, with the sign following the propagation
direction, so the velocity in medium 1 is $(p_i - p_r)/Z_1$ and in
medium 2 is $p_t/Z_2$. Continuity of pressure at $x=0$ gives
$p_i + p_r = p_t$; continuity of normal velocity gives
$(p_i - p_r)/Z_1 = p_t/Z_2$. Eliminate $p_t$: divide the second by the
first to get $(p_i - p_r)/(p_i + p_r) = Z_1/Z_2$. Solving for
$R = p_r/p_i$ yields $R = (Z_2 - Z_1)/(Z_2 + Z_1)$, and
$T = p_t/p_i = 1+R = 2Z_2/(Z_1+Z_2)$.
\end{exercise}

\begin{exercise}
Show that the natural frequencies of a rectangular room with rigid
walls are $f_{n_x, n_y, n_z} = (c/2)\sqrt{(n_x/L_x)^2 + (n_y/L_y)^2
+ (n_z/L_z)^2}$. State the boundary conditions used.
\paragraph{Solution sketch.} Seek separable standing-wave solutions of
$\partial_{tt}p = c^2\nabla^2 p$ of the form
$p = X(x)Y(y)Z(z)\cos\omega t$. Rigid walls require zero normal
particle velocity, hence zero pressure gradient normal to each wall:
$\partial_x p = 0$ at $x=0$ and $x=L_x$, and similarly in $y,z$. The
$x$ factor satisfying these Neumann conditions is
$X = \cos(n_x\pi x/L_x)$ with integer $n_x$; likewise for $y,z$.
Substituting into the Helmholtz equation $\nabla^2 p + k^2 p = 0$ gives
\[
k^2 = \left(\frac{n_x\pi}{L_x}\right)^2 +
\left(\frac{n_y\pi}{L_y}\right)^2 +
\left(\frac{n_z\pi}{L_z}\right)^2 .
\]
With $k = \omega/c = 2\pi f/c$, solving for $f$ gives the stated
formula. The boundary conditions are the rigid-wall (Neumann) pressure
conditions on all six faces.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the wavelength of the lowest audible sound ($20\,\si{
\hertz}$) and the highest audible sound ($20\,\si{\kilo\hertz}$) in
air. Estimate the smallest object that can effectively reflect each.
\paragraph{Reference estimate.} Wavelengths: $\lambda_{20} = 343/20
\approx 17\,\si{\meter}$ and $\lambda_{20k} = 343/20{,}000 \approx
17\,\si{\milli\meter}$. An object reflects a wave cleanly when its size
is comparable to or larger than the wavelength. So a low-frequency
($20\,\si{\hertz}$) wave needs an obstacle of order $\sim 17\,\si{
\meter}$, a building, to reflect strongly, which is why bass diffracts
around ordinary objects; a $20\,\si{\kilo\hertz}$ wave is reflected by
anything down to a few centimetres. The size-to-wavelength ratio is the
single governing parameter.
\end{exercise}

\begin{exercise}
Estimate the sound pressure level at $1\,\si{\meter}$ from a
loudspeaker driver radiating $1\,\si{\watt}$ acoustic power
omnidirectionally in free space. State assumptions.
\paragraph{Reference estimate.} A point source radiating power $W$
spreads it over a sphere, so the intensity at radius $r$ is
$I = W/(4\pi r^2)$. At $r=1\,\si{\meter}$,
$I = 1/(4\pi) \approx 0.080\,\si{\watt\per\meter\squared}$. The
rms pressure is $p = \sqrt{I\rho_0 c} = \sqrt{0.080\times 415}
\approx 5.8\,\si{\pascal}$. The level is
$\text{SPL} = 20\log_{10}(5.8/2\times 10^{-5}) \approx 109\,\si{
\decibel}$. The assumptions are full-sphere (omnidirectional)
radiation, free field (no reflections), and a far-field plane-wave
impedance relation. A real driver on a baffle radiates into a
half-space, adding about $3\,\si{\decibel}$. The headline: an acoustic
watt is loud, because acoustic power converts efficiently to SPL.
\end{exercise}

\begin{exercise}
Estimate the reverberation time of a typical living room of
$60\,\si{\meter\cubed}$ volume. State assumptions about absorption
coefficients of the wall, floor, and furniture surfaces.
\paragraph{Reference estimate.} A $60\,\si{\meter\cubed}$ room is
roughly $5\times4\times3\,\si{\meter}$, with a total surface of order
$S\approx 2(20) + 2(15) + 2(12) = 94\,\si{\meter\squared}$. Take a
mean absorption $\bar\alpha\approx 0.15$ (carpet and soft furniture at
$\sim 0.3$ to $0.5$, hard walls and ceiling at $\sim 0.05$, area-
weighted). Then $A = S\bar\alpha \approx 14\,\si{\meter\squared}$-sabin
and
\[
T_{60} = 0.161\,\frac{60}{14} \approx 0.7\,\si{\second},
\]
in the expected living-room range of half a second to a second.
Lowering $\bar\alpha$ to $0.1$ (a sparsely furnished hard room) raises
the estimate to $\sim 1\,\si{\second}$; the answer is sensitive to the
soft-furnishing assumption, which is the dominant absorber.
\end{exercise}

\begin{exercise}
Estimate the maximum range at which a whale's $20\,\si{\hertz}$ song
of source level $\sim 180\,\si{\decibel}$ (re $1\,\si{\micro\pascal}$
at $1\,\si{\meter}$) could be detected by a hydrophone with a
detection threshold of $80\,\si{\decibel}$ (re $1\,\si{\micro\pascal}$),
assuming spherical spreading and an attenuation of $0.01\,\si{
\decibel\per\kilo\meter}$ at $20\,\si{\hertz}$.
\paragraph{Reference estimate.} The available propagation loss is
$180 - 80 = 100\,\si{\decibel}$. Spherical spreading loss is
$20\log_{10}(r)$ with $r$ in metres, plus absorption $\alpha r$.
Trying pure spherical spreading: $100 = 20\log_{10}(r)$ gives
$r = 10^{5}\,\si{\meter} = 100\,\si{\kilo\meter}$. The absorption at
that range is only $0.01\times 100 = 1\,\si{\decibel}$, negligible, so
the answer is set by spreading alone: $r\approx 100\,\si{\kilo\meter}$.
The point of the exercise is that at $20\,\si{\hertz}$ absorption is
almost irrelevant; geometric spreading is the limiting loss, and in
the SOFAR channel where spreading is closer to cylindrical
($10\log_{10}r$) the detection range grows to thousands of kilometres.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a numerical solver for the one-dimensional acoustic wave
equation $\partial_{tt} p = c^2 \partial_{xx} p$ on a closed-tube
domain with hard-wall boundary conditions. Verify that the standing-
wave frequencies are $f_n = n c/(2L)$ for $n = 1, 2, 3$.
\paragraph{Solution sketch.} Use the explicit central-difference
leapfrog scheme $p^{n+1}_i = 2p^n_i - p^{n-1}_i + C^2(p^n_{i+1} -
2p^n_i + p^n_{i-1})$ with Courant number $C = c\,\Delta t/\Delta x \le
1$ for stability. Impose hard walls (Neumann, zero pressure gradient)
by mirroring the boundary neighbour, $p_0 = p_1$ and $p_{N} =
p_{N-1}$. Start from a broad off-centre bump with zero initial
velocity, record the pressure at a wall (a pressure antinode for every
mode), and take the FFT of the time history. The reference
implementation \texttt{code/wave\_1d.py} recovers
$f_1 = 172\,\si{\hertz}$, $f_2 = 344\,\si{\hertz}$,
$f_3 = 516\,\si{\hertz}$ for $L=1\,\si{\meter}$, $c=343\,\si{
\meter\per\second}$, matching the analytic ladder $nc/(2L) = 171.5n$ to
better than a hertz. Violating $C\le 1$ makes the scheme blow up.
\end{exercise}

\begin{exercise}
Simulate the reflection and transmission of a Gaussian pulse incident
normally on an interface between two media with impedance ratio
$Z_2/Z_1 = 4$. Verify that the reflected and transmitted pulse
amplitudes match the analytical reflection and transmission
coefficients.
\paragraph{Solution sketch.} Extend the leapfrog solver of the
previous exercise to a domain with a step change in $c$ (and hence $Z$)
at the midpoint, using the wave equation in the form
$\partial_{tt}p = c(x)^2\partial_{xx}p$ with absorbing or far ends.
Launch a Gaussian pressure pulse toward the interface. For $Z_2/Z_1 =
4$ the analytic coefficients are $R = (4-1)/(4+1) = 0.6$ and
$T = 2(4)/(1+4) = 1.6$ in pressure (note $T>1$ in pressure is allowed;
it is intensity, not pressure, that is conserved). Measure the peak
amplitude of the reflected and transmitted pulses relative to the
incident pulse; they should land near $0.6$ and $1.6$. Check energy:
$R^2 + T_I = 0.36 + 0.64 = 1$ with $T_I = 4Z_1Z_2/(Z_1+Z_2)^2 = 0.64$.
The common student error is to expect the transmitted pressure to be
smaller than the incident; into a stiffer medium it is larger.
\end{exercise}

\begin{exercise}
Compute the room-mode spectrum (the natural frequencies up to
$200\,\si{\hertz}$) of a rectangular room of $5 \times 4 \times 2.5\,
\si{\meter}$. Identify which modes are axial, tangential, and
oblique. Plot the cumulative mode count versus frequency and
extract the Schroeder frequency at which the modal density exceeds
ten modes per critical bandwidth.
\paragraph{Solution sketch.} Loop over integer triples $(n_x,n_y,n_z)$,
compute each $f_{n_x,n_y,n_z}$ from the rectangular-room formula, keep
those below $200\,\si{\hertz}$, and classify by the number of nonzero
indices (one = axial, two = tangential, three = oblique).
\texttt{code/room\_modes.py} finds sixty-five modes below
$200\,\si{\hertz}$ (eleven axial, thirty-two tangential, twenty-two
oblique), with the lowest at $34.3\,\si{\hertz}$. The cumulative count
(figure~\ref{fig:vol03ch07:room-modes}) climbs as $f^3$. The Schroeder
frequency is conventionally $f_S \approx 2000\sqrt{T_{60}/V}$; for a
$50\,\si{\meter\cubed}$ room with $T_{60}\approx 0.5\,\si{\second}$,
$f_S \approx 2000\sqrt{0.5/50} = 200\,\si{\hertz}$, the upper edge of
the modal regime computed here.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a single-stage acoustic enclosure to reduce the noise from a
$\sim 90\,\si{\decibel}$ machine at $500\,\si{\hertz}$ to an
external level of $\sim 60\,\si{\decibel}$. Specify the wall
material, the wall thickness, the seal at the door, and the
internal-absorber configuration.
\paragraph{Model-answer sketch.} The required insertion loss is
$\sim 30\,\si{\decibel}$ at $500\,\si{\hertz}$. The transmission loss
of a single panel follows the mass law, $\text{TL} \approx
20\log_{10}(m''f) - 47\,\si{\decibel}$ with $m''$ the surface mass in
$\si{\kilogram\per\meter\squared}$. For $30\,\si{\decibel}$ at
$500\,\si{\hertz}$ the panel needs $m''f \approx 10^{(30+47)/20}
\approx 7{,}000$, so $m'' \approx 14\,\si{\kilogram\per\meter\squared}$,
met by $\sim 1.8\,\si{\milli\meter}$ of steel or
$\sim 16\,\si{\milli\meter}$ of plywood with a mass-loaded barrier
layer. Line the interior with $\sim 50\,\si{\milli\meter}$ of porous
absorber to kill the enclosure's own reverberant buildup (without it
the internal level rises and erodes the insertion loss). The door is
the weak path: a single $1\,\si{\milli\meter}$ gap around its perimeter
can drop the insertion loss by ten decibels or more, so specify a
compression gasket and a continuous seal. The lesson: the enclosure is
only as good as its worst leak.
\end{exercise}

\begin{exercise}
Design a duct silencer to attenuate $\sim 30\,\si{\decibel}$ at
$1\,\si{\kilo\hertz}$ in a ventilation duct of $0.5\,\si{\meter}$
internal diameter. Specify the silencer's length and the absorber
configuration. State the pressure drop the silencer adds to the
duct.
\paragraph{Model-answer sketch.} Use a dissipative (lined) silencer.
The Sabine duct-attenuation estimate is $\Delta L \approx 1.05\,
\alpha^{1.4}\,(P/A)\,\ell$ in decibels, with $\alpha$ the lining
absorption, $P/A$ the lined-perimeter-to-cross-section ratio, and
$\ell$ the length. A $1\,\si{\kilo\hertz}$ wavelength is
$0.34\,\si{\meter}$, comparable to the duct diameter, so the lining is
effective. Split the duct with a central baffle to halve the passage
width to $\sim\lambda/2$ and double the lined perimeter; with
$\alpha\approx 0.8$ porous lining $\sim 100\,\si{\milli\meter}$ thick,
a length of $\sim 1\,\si{\meter}$ delivers $\sim 30\,\si{\decibel}$.
The pressure-drop penalty comes from the reduced free area and added
wetted surface; keep the open area above $\sim 50\,\%$ and the face
velocity modest ($\lesssim 10\,\si{\meter\per\second}$) to hold the
static-pressure loss to tens of pascals, and protect the porous lining
with a perforated facing and a thin film so the airstream does not
erode the fibres.
\end{exercise}

\begin{exercise}
Design a Helmholtz resonator to absorb sound at $250\,\si{\hertz}$
in a small treated room. Specify the resonator's cavity volume, the
neck cross-section, and the neck length.
\paragraph{Model-answer sketch.} The Helmholtz resonance is
$f_0 = \frac{c}{2\pi}\sqrt{S/(V\,L_{\text{eff}})}$, with $S$ the neck
cross-section, $V$ the cavity volume, and $L_{\text{eff}} = L + 1.7a$
the neck length plus end corrections ($a$ the neck radius). Pick a neck
of radius $a = 25\,\si{\milli\meter}$ ($S = \pi a^2 \approx 2.0\times
10^{-3}\,\si{\meter\squared}$) and physical length $L = 50\,\si{
\milli\meter}$, so $L_{\text{eff}} \approx 0.05 + 1.7(0.025) =
0.093\,\si{\meter}$. Solve for the cavity volume:
\[
V = \frac{S c^2}{(2\pi f_0)^2 L_{\text{eff}}}
= \frac{2.0\times 10^{-3}(343)^2}{(2\pi\cdot 250)^2(0.093)}
\approx 1.0\times 10^{-3}\,\si{\meter\cubed},
\]
about one litre. Add a little porous damping in the neck to broaden
the (otherwise sharp) absorption band; an undamped resonator absorbs
strongly only within a few hertz of $f_0$. Tune in practice by
adjusting the neck length, which moves $f_0$ without rebuilding the
cavity.
\end{exercise}

\begin{exercise}
Design an underwater hydrophone for the band $1$ to $10\,\si{\kilo
\hertz}$, specifying the active material, the transducer geometry,
the matching layer, and the sensitivity (output voltage per unit
pressure).
\paragraph{Model-answer sketch.} Use a piezoelectric ceramic (PZT) in
a configuration whose lowest resonance sits well above the band, so the
device responds flat across $1$ to $10\,\si{\kilo\hertz}$; a small
cylinder or sphere of a few centimetres has its resonance in the tens
of kilohertz, satisfying this. Below resonance the sensitivity is set
by the piezoelectric voltage constant $g$ and the element thickness,
giving open-circuit sensitivities of order $-200$ to $-180\,\si{
\decibel}$ re $1\,\si{\volt}/\si{\micro\pascal}$, that is, roughly a
microvolt to tens of microvolts per pascal, so a preamplifier mounted
at the element is needed before cable capacitance loads the signal.
The hydrophone radiates into water, whose impedance ($1.5\times
10^{6}\,\text{rayl}$) is far closer to PZT ($\sim 33\times 10^{6}$)
than air is, so a single quarter-wave matching layer of $Z_m =
\sqrt{Z_t Z_{\text{water}}} \approx 7\times 10^{6}\,\text{rayl}$ (a
metal-loaded polymer) suffices to flatten the response; a waterproof,
acoustically transparent boot (an oil-filled polyurethane) couples the
element to the sea while excluding water from the electronics.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A room exhibits a strong audible peak at $\sim 100\,\si{\hertz}$
when bass-frequency music is played. Identify the most probable
mechanism, the diagnostic measurement, and the corrective action.
\paragraph{Solution sketch.} The probable mechanism is a room mode: a
standing wave whose frequency falls near $100\,\si{\hertz}$ and whose
pressure antinode lies near the listening or measurement position. For
a domestic room a dimension of $\sim 1.7\,\si{\meter}$ (a half-
wavelength at $100\,\si{\hertz}$) or an integer multiple of it would
place an axial mode there. The diagnostic measurement is a swept-sine
or pink-noise excitation with the microphone moved around the room: a
true mode shows a pressure peak that grows toward the room boundaries
and corners and nulls at the mode's nodal plane, distinguishing it
from an electronic or loudspeaker resonance, which would not move with
position. The correction is bass trapping (a porous or membrane
absorber placed at a pressure antinode, that is, in a corner where all
modes peak), repositioning the listener or the loudspeaker off the
antinode, or parametric equalisation to notch the offending frequency.
\end{exercise}

\begin{exercise}
A jet engine's exhaust impinges on an adjacent fuselage panel. The
panel develops visible cracks at a fastener row after several
hundred operating hours. Identify the failure mechanism, the
calculation a design review should have performed, and the
inspection that would have caught the precursor.
\paragraph{Solution sketch.} The mechanism is acoustic-induced
fatigue: the high-intensity exhaust noise (often above
$150\,\si{\decibel}$) excites the panel at its structural natural
frequencies, the resulting cyclic stress concentrates at the fastener
row (a stiffness discontinuity and a stress raiser), and cumulative
fatigue cycles drive crack initiation and growth in hundreds of hours.
The design-stage calculation is the chain of section~7.8: measure or
estimate the SPL spectrum at the panel, compute the panel's modal
natural frequencies and the response amplitude where the excitation
spectrum overlaps a mode, convert the modal displacement to cyclic
stress at the fastener holes, and apply a fatigue-life method (S-N or
fracture-mechanics) to the resulting stress spectrum. The precursor-
catching inspection is scheduled eddy-current or dye-penetrant
inspection of the fastener row at intervals set by the predicted
crack-growth life, looking for sub-critical cracks before they reach
the visible (and near-critical) size.
\end{exercise}

\begin{exercise}
An industrial gearbox emits a tonal whine at $\sim 1\,\si{\kilo
\hertz}$ that has grown over six months. Identify the most probable
mechanism (mesh frequency tone, bearing fault, shaft alignment) and
the diagnostic measurement.
\paragraph{Solution sketch.} A growing tonal whine at a fixed
frequency points to a gear-mesh tone: the gear-mesh frequency is the
shaft rotation rate times the number of teeth, and a tone there that
grows over months signals progressing tooth wear, pitting, or a
profile error increasing the dynamic transmission error. A bearing
fault would more often appear as broadband or as a non-synchronous
tone at a bearing defect frequency (ball-pass outer/inner race),
non-integer with shaft rate; misalignment shows as elevated
once-per-revolution and twice-per-revolution vibration. The diagnostic
measurement is an order-tracked vibration or acoustic spectrum
referenced to shaft speed: compute the gear-mesh frequency from the
geometry, look for a peak there with growing sidebands spaced at the
shaft rate (the wear signature), and trend the amplitude over time
against a baseline. Confirm with an oil-debris analysis (metal
particles from tooth-surface wear).
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Identify a real or hypothetical case in which high-intensity
acoustic loading (from a jet engine, a rocket motor, a turbomachinery
blade-pass, or aerodynamic buffet) caused fatigue failure of a
structural panel. State, in $200$ to $300$ words, the design-stage
calculation that should have been performed and the inspection
discipline that should have caught the precursor.
\paragraph{Model-answer sketch.} A defensible hypothetical: a rocket
launch vehicle's payload-fairing acoustic environment exceeds
$140\,\si{\decibel}$ during the first seconds of flight (reflected
engine noise off the pad and aerodynamic buffet through transonic),
and a thin skin panel near a frame attachment cracks at the rivet line.
A strong answer names the loading source, the panel mode excited, and
the stress-raiser location, then states the design-stage chain: define
the acoustic-loading spectrum from a standard envelope (a NASA or
military specification), compute the panel modal frequencies and the
overlap with the excitation, estimate the modal stress at the rivet
holes, and apply a fatigue or fracture-mechanics life calculation with
an appropriate scatter factor. The inspection discipline is
non-destructive inspection of the fastener rows (eddy current or
penetrant) at intervals shorter than the predicted crack-growth life,
plus a fleet-leader teardown to validate the life prediction. The
graded distinction is whether the answer connects the acoustic
spectrum to a specific structural mode and a specific cyclic stress,
rather than asserting that loud noise causes cracks; the chapter's
point is that the loading is computable and the failure preventable.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer asks whether the A-weighting of sound levels is
arbitrary, since it's based on the human ear's response rather than
on physical principles. Write a one-paragraph reply that takes the
question seriously and states when A-weighting is appropriate and
when it is not.
\paragraph{Model-answer sketch.} A strong reply grants the premise:
A-weighting is not a law of physics, it is a fitted approximation to
the equal-loudness contour at moderate levels, standardised so that
measurements made anywhere are comparable. That is exactly its value
when the quantity of interest is human response: hearing-damage risk,
annoyance, speech interference. It is the right tool for occupational
and environmental noise limits because those limits are about people.
It is the wrong tool when the quantity of interest is the physical
field or the structural load: acoustic fatigue of a panel depends on
the unweighted pressure spectrum at the panel, not on what a human
would hear; low-frequency content that A-weighting discards may
dominate the structural excitation. The discipline is to weight by the
receiver's actual sensitivity, which is the ear for people and the
structural transfer function for hardware. The answer should resist
both errors: treating A-weighting as physics, and dismissing it as
arbitrary; it is a deliberate, documented receiver model.
\end{exercise}

\begin{exercise}
A senior colleague argues that noise control should always intervene
at the source, never on the path or at the receiver, on the grounds
that source control is the most effective. Write a one-paragraph
reply that addresses the cost-benefit trade-off.
\paragraph{Model-answer sketch.} The colleague is right about the
hierarchy and wrong about the absolute. Source control is the most
effective and the most durable, because it removes the energy before
it propagates and protects everyone downstream without ongoing
discipline; the source-path-receiver order is the correct default. But
effectiveness per dollar is not the same as effectiveness, and source
control is often the most expensive or technically impossible
intervention: a quieter machine may not exist, redesigning it may cost
more than the noise problem is worth, and a legacy installation cannot
be re-sourced. A sound reply states that the engineer should attack
the source first where the cost is justified, fall to path control
(enclosures, barriers, silencers) where the source is fixed, and treat
the receiver (hearing protection, relocation) as the last line, which
is cheap but fragile because it depends on consistent use. The right
answer is usually a combination chosen by cost-benefit at the dominant
path, not a doctrine. Spending on a path treatment that yields
$15\,\si{\decibel}$ for a tenth the cost of a $20\,\si{\decibel}$
source fix is the better engineering decision.
\end{exercise}

\begin{exercise}
Identify an acoustic environment in your immediate surroundings
(your bedroom, an office, a public space) whose reverberation time
and modal structure you have never measured. State, in $200$ to
$300$ words, the dominant modes you would expect from the geometry,
the dominant absorbers, and whether the environment is suited to
its use.
\paragraph{Model-answer sketch.} A complete answer measures or paces
the room's three dimensions, computes the lowest axial modes from
$f = c/(2L)$ along each, and flags any near-coincidences (dimensions
in small-integer ratios that pile modes together and produce a boomy
or uneven bass). It then estimates $T_{60}$ from the Sabine formula
using the volume and a reasoned area-weighted absorption, naming the
dominant absorbers (carpet, curtains, soft furniture, people) and the
dominant reflectors (bare walls, glass, hard floors). The judgment
step compares the estimated $T_{60}$ to the use: a bedroom or office
wants a short reverberation ($\sim 0.3$ to $0.5\,\si{\second}$) for
speech clarity and quiet, while a music room tolerates more; a glass-
and-tile space with $T_{60}$ over a second is poorly suited to
conversation. The graded distinction is whether the writer connects a
specific geometric feature to a specific audible consequence and
proposes a proportionate change (a rug, curtains, a bookshelf as a
diffuser), rather than offering generic observations. The exercise
rewards the habit of hearing a room as a measurable system.
\end{exercise}
