\chapter{Fluid statics}
\label{vol03:ch10}

\epigraph{Any solid lighter than a fluid will, if placed in the
fluid, be so far immersed that the weight of the solid will be
equal to the weight of the fluid displaced.}{Archimedes,
\emph{On Floating Bodies}, Book I, Proposition 5, in the Heath
edition \cite[Book I, Prop.~5]{hist:archimedes-floating}.}

\chapmeta{Half-life: durable. Archetypes: balance, scaling. Exercise target: 30. Project track: instrumentation and field measurement. Hazard class: none. Reader path default: standard, with sections explicitly tagged. Named cases: none directly in chapter prose; the failure section discusses dam-breach mechanisms and floating-stability loss as failure patterns without citing a specific event in the registry.}

\noindent
A fluid at rest cannot sustain a shear stress: a fluid is a
substance that deforms continuously under any applied shear,
however small. The consequence is that the only stress a fluid at
rest sustains is pressure, a scalar that varies with position
according to the body forces and the fluid's own weight. The
chapter develops the pressure field for the hydrostatic case
(gravity is the only body force), the working tools that exploit
it (manometers, pressure-on-surface calculations, buoyancy), the
extension to non-inertial frames (a fluid in an accelerating
tank, a rotating bucket), the additional physics that surface
tension contributes at small scales, and the failure modes by
which a fluid-statics design can go wrong (dam-breach, floating
instability, surface-tension-driven leakage). Three end-to-end
case studies carry the separate results to a verdict on real
structures: a concrete gravity dam that must not tip or slide, a
loaded work pontoon that must stay upright, and a pair of canal
lock gates that turn the head they hold back into the thrust that
seals them.

\section{Pressure as a scalar field}\pathtag{core}

\engterm{Pressure} at a point in a fluid is the force per unit
area exerted by the fluid on a surface element at that point.
Pressure has units of pascals, $1\,\text{Pa} = 1\,\text{N}/\text{m}
^2$. Working units in engineering practice: kilopascals, bars
($1\,\text{bar} = 10^5\,\text{Pa}$), atmospheres ($1\,\text{atm}
= 101.325\,\text{kPa}$), millimetres of mercury ($1\,\text{mmHg}
\approx 133.32\,\text{Pa}$), metres of water column ($1\,\text{m
H}_2\text{O} \approx 9.81\,\text{kPa}$), and pounds per square
inch ($1\,\text{psi} \approx 6.895\,\text{kPa}$). The variety of
units reflects the variety of measurement instruments and
historical conventions; the SI unit pascal is the working choice
for analytical work.

In a fluid at rest, pressure is a scalar function of position:
$p(x, y, z)$. The proof, due to Euler and presented in any modern
fluid-mechanics text \cite[ch.~2]{text:white-fluid-mechanics},
takes a small tetrahedral fluid element with three orthogonal
faces and one inclined face, applies force balance, and lets the
size of the tetrahedron shrink to zero. The pressures on the
three orthogonal faces appear in the force balance with
coefficients proportional to the corresponding face areas; the
inclined face's pressure appears with a coefficient proportional
to its area; the gravitational body force scales with the
tetrahedron's volume, which vanishes faster than the surface
forces. In the limit, the inclined-face pressure equals each
orthogonal-face pressure, regardless of the orientation of the
inclined face. The pressure at a point is therefore the same in
all directions: $p$ is a scalar.

The same argument fails for a moving fluid with viscous shear, or
for a fluid in motion with significant inertia: the surface
stresses then include shear components that survive the
limit. The pressure-is-a-scalar result is specific to fluids at
rest (or, more generally, to fluids in which the only stress
component on any surface is the normal compression).

\engterm{Absolute and gauge pressure}. The pressure at a point is
typically reported as one of two quantities:

\begin{itemize}[leftmargin=*]
  \item \engterm{Absolute pressure}: the pressure measured from a
    reference of perfect vacuum. The absolute pressure is what
    appears in thermodynamic equations of state (the ideal gas
    law) and in compressible-flow analyses.
  \item \engterm{Gauge pressure}: the pressure measured from a
    reference of local atmospheric pressure. Gauge pressure is
    what a typical pressure-measurement instrument reports
    directly: a Bourdon-tube gauge at atmospheric reads zero;
    an open-end manometer at atmospheric reads zero. Gauge
    pressure is the working choice for most engineering
    calculations of structural and hydraulic loads.
\end{itemize}

The relationship is $p_{\text{abs}} = p_{\text{gauge}} + p_{
\text{atm}}$, with $p_{\text{atm}} \approx 101\,\text{kPa}$ at
sea level. A pressure below atmospheric is reported as a positive
\engterm{vacuum} (or \engterm{negative gauge pressure}): a
vacuum of $50\,\text{kPa}$ corresponds to an absolute pressure
of $\sim 51\,\text{kPa}$. The convention is recorded explicitly
in every working calculation; mixing the two is a routine source
of errors in pressure-vessel design.

\begin{archetype}[Balance]
The archetype of fluid statics is \emph{balance}: every fluid
element is in static equilibrium under gravity and the pressure
gradient. Two of the most consequential engineering quantities,
the buoyancy of a floating body and the hydrostatic force on a
submerged wall, are direct integrations of this elementary
balance. The pleasure of fluid statics is that the same balance
equation, written for a fluid element of arbitrary size and
shape, gives both the wine in the cup and the ocean against the
hull of a ship. The discipline is to identify the element, write
the balance, and integrate; the discipline is identical whether
the fluid is water at room temperature, magma in a planetary
mantle, or compressed gas in a vessel.
\end{archetype}

\section{Hydrostatic pressure}\pathtag{core}

Consider a small fluid element of volume $dV = dx\,dy\,dz$ in a
gravitational field with acceleration $\vec{g} = -g\hat{k}$
(pointing downward, $z$ pointing up). The element is in static
equilibrium under the pressure forces on its six faces and its
own weight. The net pressure force in the $x$ direction is
$-(\partial p/\partial x)\,dV$ (the pressure on the $+x$ face
minus the pressure on the $-x$ face, times the face area); the
same expression with $y$ replaces $x$ for the $y$ direction; the
$z$-direction net pressure force is $-(\partial p/\partial z)\,
dV$. Adding the gravitational body force $-\rho g \,dV$ in the
$z$ direction and setting the sum to zero gives the
\engterm{hydrostatic equation}:
\[
\nabla p = \rho \vec{g}, \quad \text{or, in components,} \quad
\frac{\partial p}{\partial x} = 0, \quad \frac{\partial p}{
\partial y} = 0, \quad \frac{\partial p}{\partial z} = -\rho g.
\]
The pressure is constant in any horizontal plane (the first two
components) and varies linearly with depth (the third). The
result is what Pascal stated in 1663 in the \emph{Trait{\'e} de
l'{\'e}quilibre des liqueurs} \cite[Treatise I, ch.~1]{
hist:pascal1663}: a fluid at rest transmits pressure equally in
all directions, and the pressure increases with depth at a rate
set by the fluid's specific weight $\gamma = \rho g$.

For an incompressible fluid (a working approximation for liquids
under engineering pressures: water's compressibility produces a
density change of less than $0.5\,\%$ across a $1\,\text{km}$
depth), the third equation integrates trivially:
\[
p(z) = p_0 - \rho g (z - z_0) = p_0 + \rho g h,
\]
where $h = z_0 - z$ is the depth below the reference level $z_0$
at which the pressure is $p_0$. The result is the
\engterm{hydrostatic depth formula}: pressure increases linearly
with depth, at a rate $\rho g$ per unit depth. Working numbers
for water ($\rho \approx 10^3\,\si{\kilogram\per\meter\cubed}$,
$g \approx 9.81\,\text{m/s}^2$, $\rho g \approx 9.81\,\text{kPa/m}$),
the pressure increase per metre is approximately $10\,\text{kPa}$;
a $10\,\si{\meter}$ water column adds approximately one
atmosphere of pressure.

\engterm{Pascal's principle}. The hydrostatic equation has an
immediate consequence: a pressure change applied at any point in
a confined incompressible fluid transmits to every other point
without diminution (apart from the static depth-dependence). The
result is the basis of the hydraulic press: a small force on a
small piston produces a large pressure, transmitted through the
fluid to a large piston where it produces a large force, with the
ratio of forces equal to the ratio of piston areas. Pascal stated
the principle and demonstrated it experimentally in 1663 with the
hydraulic-press apparatus he reports in the second treatise
\cite[Treatise II]{hist:pascal1663}.

Two features of the transmission bear on how a hydraulic machine is
sized. The pressure the small piston raises adds to, and does not replace,
the static depth term, so a working press carries $p(z)=p_{\text{applied}}
+\rho g h$ at every point and a designer who ignores the applied term when
sizing a seal at the bottom of a tall ram underestimates the load by the
whole pumped pressure. The transmission is also loss-free only in the
static idealisation: the incompressible fluid stores no energy, so the
work in at the small piston equals the work out at the large one, force
times distance conserved. The large piston moves as far less than the
small one as its area is larger, which is the exchange rate a hydraulic
jack pays and a worked example later in the chapter carries through in
full. Real fluids depart from the ideal through a small compressibility, a
line elasticity that swells under pressure, and any trapped gas, all of
which store energy elastically and make the machine feel spongy; bleeding
the air from a hydraulic line is the practical cure, because a few per
cent of gas by volume dominates the compliance of an otherwise stiff
liquid column.

\begin{mastery}
\textbf{The hydrostatic paradox: pressure on the base does not know the
shape of the vessel.} Fill three open vessels to the same height $H$ with
the same liquid: a straight cylinder, a vessel that flares outward toward
the top, and one that narrows toward the top. The pressure at the base is
$\rho g H$ in every case, because the hydrostatic law depends on depth
alone and not on how much liquid stands above. The downward force on a
base of area $A$ is therefore $\rho g H A$ in all three, even though the
flared vessel holds far more liquid than the narrow one and the narrow one
holds far less than a straight cylinder of the same base. This is Pascal's
hydrostatic paradox, and the resolution is that the walls carry the
difference. In the flared vessel the sloping walls push down on the liquid
with a vertical component, and the base feels only $\rho g H A$, less than
the full weight of liquid; in the narrowing vessel the walls pull up on
the liquid (they push in along their inward-leaning normals, with an
upward component), and the base feels $\rho g H A$, more than the weight of
the liquid it holds. The weight of the liquid is always carried by the
whole wetted boundary, base and walls together, and only the base share
follows the simple depth law. The lesson sizes real vessels: the base
plate of a conical hopper, a flared settling tank, or a stepped standpipe
is loaded by its liquid height, not by the tonnage it stores, and reading
the load off the stored mass rather than the head is the error the paradox
exists to prevent.
\end{mastery}

\engterm{The compressibility correction}. For a compressible
fluid (a gas, or a liquid at extreme depth), the density depends
on the pressure: $\rho = \rho(p)$. The hydrostatic equation then
reads $dp/dz = -\rho(p) g$, a first-order ODE for $p(z)$. For an
ideal gas at uniform temperature $T$, $\rho = p M /(R T)$ (with
$M$ the molar mass and $R$ the universal gas constant), and the
integration gives the \engterm{barometric formula}:
\[
p(z) = p_0 \exp\!\left(-\frac{M g (z - z_0)}{R T}\right).
\]
The pressure decreases exponentially with height. The scale height
$H = RT/(Mg)$ for the Earth's atmosphere at $T = 288\,\text{K}$
is about $8.4\,\si{\kilo\meter}$; the pressure halves every
$\sim 5.5\,\si{\kilo\meter}$ of altitude. Section 10.6 extends
the barometric model to the standard temperature-gradient
atmosphere of meteorology.

\begin{principle}[Pressure rises linearly with depth in a static
incompressible fluid]
For an incompressible fluid at rest under uniform gravity, the
pressure at depth $h$ below a free surface where the pressure is
$p_0$ is $p(h) = p_0 + \rho g h$. The result is exact within the
incompressibility assumption and is the working equation for
every problem of pressure on a submerged surface, every manometer
reading, and every hydrostatic load on a dam, hull, or tank wall.
The single rule, $\rho g$ per unit depth, carries most of fluid
statics.
\end{principle}

\input{volumes/03-force/ch10-fluid-statics/figures/fig-pressure-depth.tex}

\section{Manometers}\pathtag{standard}

A \engterm{manometer} is an instrument for measuring pressure
by reading the height of a fluid column. The instrument exists in
several configurations, each with its own working formula derived
from the hydrostatic depth equation.

\engterm{The open-end U-tube manometer}. A U-shaped tube
partially filled with a manometer fluid (mercury, water, or oil)
is connected at one leg to the pressure source and at the other
leg open to the atmosphere. The manometer fluid stands at
different heights in the two legs; the difference $\Delta h$
measures the gauge pressure at the connection point. Reading the
hydrostatic equation along a fluid path from the connection point
to the open atmospheric leg (descending through the manometer
fluid and rising through air),
\[
p_{\text{conn}} + \rho_{\text{manom}} g \Delta h - \rho_{\text{air
}} g \Delta h_{\text{air}} = p_{\text{atm}}.
\]
With $\rho_{\text{air}} \ll \rho_{\text{manom}}$, the air column
term is negligible, and the manometer reads $p_{\text{conn,gauge}}
= \rho_{\text{manom}} g \Delta h$. For mercury ($\rho \approx
13{,}600\,\si{\kilogram\per\meter\cubed}$), each millimetre of
$\Delta h$ corresponds to $\approx 133\,\text{Pa}$; for water,
each millimetre corresponds to $\approx 9.81\,\text{Pa}$.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-manometer.tex}

The path-tracing reduction, including the small process-fluid column
that the approximation drops, is carried out in
\texttt{code/manometer.py}; for a $120\,\si{\milli\meter}$ mercury
reading on a gas line the air column changes the answer by less than
$0.01\,\%$, which is the quantitative justification for neglecting it.

\engterm{The differential manometer}. A U-tube connected at both
legs to two different pressure sources reads the pressure
difference directly. The working equation, with $\rho_{\text{manom}
}$ the manometer fluid and $\rho_{\text{process}}$ the process
fluid in the connecting lines,
\[
p_1 - p_2 = (\rho_{\text{manom}} - \rho_{\text{process}}) g \Delta
h,
\]
where $\Delta h$ is the height difference of the manometer fluid
between the two legs. The process-fluid density appears with a
minus sign because the process fluid fills the connecting lines
above the manometer fluid on both sides; the difference in the
process-fluid column heights cancels in the equation.

\engterm{The inclined manometer}. A manometer whose legs are
inclined at an angle $\theta$ to the horizontal amplifies the
displacement of the manometer fluid by a factor of $1/\sin\theta$
compared to a vertical manometer. An inclined manometer at $\theta
= 5^{\circ}$ produces a fluid-column displacement of $\sim 11.5$
times the vertical equivalent, allowing readings of small
pressures to a higher resolution.

\engterm{The Bourdon-tube gauge}. A coiled or flattened metal
tube responds to internal pressure by partially uncoiling; the
displacement is mechanically transmitted to a pointer on a dial.
The gauge measures gauge pressure (zero at atmospheric) by
direct mechanical action, with a typical accuracy of $\pm 0.5\,\%$
to $\pm 1\,\%$ of full scale. The Bourdon gauge does not read at
the high accuracy of a precision manometer but is robust,
compact, and useful in field measurement.

\engterm{Electronic pressure transducers}. A diaphragm-based
sensor with a strain-gauge or piezoresistive element converts
pressure into an electrical signal. Modern silicon-piezoresistive
transducers achieve accuracies of $\pm 0.05\,\%$ to $\pm 0.25\,\%$
of full scale (current as of 2025) over a pressure range from
vacuum to thousands of bar. The transducer is the working
instrument for most industrial and laboratory pressure
measurement.

The choice among these instruments turns on a resolution-and-range
trade that the hydrostatic law makes quantitative. A liquid-column
manometer converts pressure into a length through the single factor
$\rho g$, so its resolution is set by how finely a height can be read
and its range by how tall a column the instrument can hold. A water
column reads $9.81\,\text{Pa}$ per millimetre and a mercury column
$133\,\text{Pa}$ per millimetre, so water gives fourteen times the
resolution of mercury for the same height reading but needs fourteen
times the column height to reach the same pressure. The inclined
manometer buys resolution the same way, stretching a small vertical rise
into a long slope reading by the factor $1/\sin\theta$, at the cost of a
long tube and a growing sensitivity to levelling error, because a tube
off level by a small angle adds a bias that also scales as $1/\sin\theta$.
The well-type manometer, worked in a later example, trades one readable
leg for a large reservoir and carries a small computable area-ratio bias
in return. Against all of these the electronic transducer gives up the
direct traceability of a height read against gravity and a known density,
and gains range, speed, and an electrical output, at the price of a
calibration that drifts and must be checked back against a column or a
deadweight tester. The discipline that unifies the family is that every
one of them is the hydrostatic law read backward, a pressure inferred
from a displacement, and every one carries an uncertainty that traces to
the density, the local gravity, and the length or displacement it
actually measures.

The manometer-based diagnostics are the engineer's standard
hydrostatic reasoning tool. The discipline is to trace the
hydrostatic path from one terminal of the instrument to the
other, accumulating $\rho g \Delta z$ contributions at each
fluid column, with the sign set by whether the path descends
(increasing pressure) or rises (decreasing pressure). A manometer
reading is wrong by a factor or a sign whenever the path tracing
is skipped; the discipline is to draw the manometer and trace the
path on every problem.

\section{Forces on submerged surfaces}\pathtag{core}

A submerged surface (a dam, a tank wall, a hull plate, a
sluice gate) is loaded by the hydrostatic pressure field on its
wetted side. The total force and the moment that the force
produces about a chosen axis are the working quantities for the
structural design of the surface.

For a flat submerged plate of arbitrary shape, oriented at an
arbitrary angle to the vertical, the resultant hydrostatic force
on one side and its line of action are calculated by integration
of the hydrostatic pressure over the wetted area.

\engterm{Resultant force on a flat submerged surface}. The
hydrostatic pressure varies linearly with depth: $p(h) = \rho g h$
(gauge). The resultant force on a surface of area $A$ is
\[
F = \int_A p \,dA = \int_A \rho g h \,dA = \rho g \int_A h \,dA
= \rho g h_C A,
\]
where $h_C = \frac{1}{A}\int_A h\,dA$ is the depth of the surface's
\engterm{centroid} below the free surface. The force is therefore
\[
F = \rho g h_C A = p_C A,
\]
with $p_C = \rho g h_C$ the pressure at the centroid. The
working formula is striking in its simplicity: \emph{the total
hydrostatic force on a flat submerged plate equals the pressure
at the centroid, times the area}. The result is independent of
the plate's shape and the angle of inclination. It is one of the
most useful results in fluid statics.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-pressure-prism.tex}

\engterm{Line of action: the centre of pressure}. The line of
action of the resultant force is not through the centroid; the
pressure increases with depth, so the line of action is below
the centroid. The \engterm{centre of pressure}, where the
resultant acts, is at depth
\[
h_P = h_C + \frac{I_{xx,C}}{h_C A},
\]
where $I_{xx,C}$ is the second moment of the area about the
horizontal axis through the centroid (perpendicular to the
direction of increasing depth). The shift below the centroid,
$I_{xx,C}/(h_C A)$, is the second-moment correction. For a
rectangular vertical plate of height $b$ submerged with its top
edge at depth $h_T$, the centroid is at depth $h_C = h_T + b/2$,
$I_{xx,C} = b^3/12 \cdot \text{(width)}$, and the centre of
pressure is at depth $h_P = h_C + b^2/[12(h_T + b/2)]$.

The shift toward the deeper edge is most pronounced when the
plate's depth is comparable to the depth of the centroid (a
plate at the free surface, $h_T = 0$): the centre of pressure
is then at $h_P = 2b/3$ from the surface, i.e., at the lower
third-point of the plate. The shift becomes small when the
plate's depth is small compared to the depth of the centroid (a
small plate at great depth): the pressure field is then nearly
uniform across the plate, and the centre of pressure approaches
the centroid.

The offset $I_{xx,C}/(h_C A)$ deserves a second look, because its
scaling settles where a gate's hinge and framing carry the load. Write
the second moment as $I_{xx,C} = A\,k^2$, with $k$ the radius of
gyration of the area about its centroidal axis; $k$ is a length that
measures how far the area spreads from its own centre, and it grows in
proportion to the plate's own size. The offset is then $k^2/h_C$, so it
grows as the square of the plate size and falls as the first power of the
centroidal depth. Two plates of the same shape at the same depth carry
their resultants at the same fractional position; a plate scaled up by a
factor of two throws its resultant four times further below the centroid.
The dependence is why a tall reservoir gate near the surface must frame
for a resultant sitting far down its face, while a small inspection port
on a deep pipe can be treated as loaded through its centre with negligible
error. Figure~\ref{fig:vol03ch10:cp-shift} plots the offset against depth
for three plate sizes and shows the inverse-depth decay directly: the
offset that reaches half a metre for a large plate near the surface has
shrunk to a few centimetres by the time the same plate sits several metres
down.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-cp-shift.tex}

\engterm{Resultant force on a curved submerged surface}. A
curved surface (a cylindrical tank wall section, a hull plate
with curvature, a curved gate) is analysed by decomposing the
hydrostatic force into horizontal and vertical components. The
horizontal component on the curved surface equals the
hydrostatic force on the surface's projection onto a vertical
plane (the same flat-plate calculation as above, applied to the
projection). The vertical component equals the weight of the
fluid column directly above (or below) the surface, between the
surface and the free surface. The two components combine
vectorially to give the resultant magnitude and direction.

The decomposition into horizontal and vertical components is
exact and is the working tool for curved surfaces. It avoids the
direct surface-integral over the curved surface, which is harder
to set up correctly.

\begin{mastery}
\textbf{Why the component method works, and how buoyancy falls out of
it.} The hydrostatic force on a surface element $d\vec{A}$ is $-p\,d\vec{A}$,
with the area vector pointing along the outward normal. Resolve the
element into its projections onto the three coordinate planes. The
horizontal force in the $x$ direction is
\[
F_x = -\int_S p\, (d\vec{A})_x = -\int_{S_x} p\,dA_x,
\]
where $S_x$ is the projection of the curved surface onto the plane
perpendicular to $x$, a flat vertical area. The integral over $S_x$ is
exactly the flat-plate calculation $p_C A$ applied to that projection,
which is the rule stated above: the horizontal component on a curved
surface equals the force on its vertical projection. The vertical
component is
\[
F_z = -\int_S p\, (d\vec{A})_z.
\]
Write $p = \rho g h$ with $h$ the depth below the free surface, and read
$dA_z$ as the horizontal footprint of each element. Each vertical strip
contributes $\rho g h\,dA_z$, which is the weight of a fluid column of
height $h$ over the footprint $dA_z$. Summing the strips gives the weight
of the fluid volume bounded above by the free surface and below by the
curved surface, the result stated for the vertical component. When the
fluid lies below the surface (a curved surface with water beneath it, like
the bottom of a floating hull), the same integral counts the column from
the surface down to the surface and the sign flips, giving an upward
force equal to the weight of the displaced volume.

Now close the surface. For a fully submerged body the curved surface
wraps all the way around. The horizontal components from opposite sides
cancel, because the two vertical projections are equal and oppositely
directed. The vertical components do not cancel: the upward push on the
underside counts the weight of fluid from the free surface down to the
underside, the downward push on the topside counts the weight from the
free surface down to the topside, and the difference is the weight of the
fluid between the topside and the underside, which is exactly the volume
of the body. The net force is therefore $\rho_f g V$ upward, Archimedes'
principle, recovered as a special case of the component method with no
new physics. Buoyancy is not a separate law; it is what the curved-surface
decomposition gives when the surface closes on itself.
\end{mastery}

\subsection{Worked example: a tilted gate}

A rectangular gate of width $w = 2\,\si{\meter}$ and height $b =
3\,\si{\meter}$ is hinged at its top edge to a wall of a water
reservoir. The gate is inclined at $\theta = 60^{\circ}$ from
horizontal, with its top edge $h_T = 1\,\si{\meter}$ below the
water surface. Find the resultant hydrostatic force on the gate
and the moment about the hinge.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-gate.tex}

The depth of the gate's centroid (along the gate, at distance
$b/2$ from the top edge) below the free surface is
\[
h_C = h_T + (b/2) \sin\theta = 1 + 1.5 \times \sin 60^{\circ}
= 1 + 1.299 \approx 2.30\,\si{\meter}.
\]
The pressure at the centroid is $p_C = \rho g h_C \approx 10^3
\times 9.81 \times 2.30 \approx 22.6\,\text{kPa}$. The resultant
force on the gate (acting perpendicular to the gate surface) is
\[
F = p_C A = 22.6 \times 10^3 \times (2 \times 3) \approx 1.35
\times 10^5\,\text{N} = 135\,\text{kN}.
\]
The line of action (centre of pressure, along the gate measured
from the top edge) is at
\[
s_P = (b/2) + \frac{I_{xx,C} \sin\theta}{h_C A}
= 1.5 + \frac{(w b^3/12) \sin\theta}{h_C A}
= 1.5 + \frac{(2 \times 27/12) \times 0.866}{2.30 \times 6}
\approx 1.5 + 0.282 \approx 1.78\,\si{\meter}.
\]
The hinge moment, taking the hinge axis at the top edge,
\[
M_{\text{hinge}} = F \times s_P \approx 1.35 \times 10^5 \times
1.78 \approx 2.40 \times 10^5\,\text{N}\cdot\text{m} \approx
240\,\text{kN}\cdot\text{m}.
\]
The hinge and its mounting must carry this moment under the
service load, with appropriate safety factors. The same calculation
is reproduced numerically in \texttt{code/submerged\_surface.py}, which
also runs the vertical-dam case of the exercises; the closed-form and
the numerical answers agree to the rounding shown.

\section{Buoyancy and stability of floating bodies}\pathtag{core}

A body partially or fully submerged in a fluid is acted on by a
buoyancy force equal to the weight of the fluid the body
displaces. The result is the \engterm{principle of Archimedes},
first stated in \emph{On Floating Bodies}, Book I, propositions
$5$ through $7$ \cite[Book I, Props.~5--7]{
hist:archimedes-floating}. In modern notation: a body of volume
$V_{\text{disp}}$ submerged in a fluid of density $\rho_f$
experiences a buoyancy force
\[
F_B = \rho_f g V_{\text{disp}},
\]
directed vertically upward and acting through the
\engterm{centre of buoyancy}, the centroid of the displaced fluid
volume. The proof is a direct application of the hydrostatic
pressure distribution: the net upward force on a submerged body
is the integral of the pressure difference between the bottom
and top of the body over the body's vertical projection, which
equals $\rho_f g V_{\text{disp}}$ exactly.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-buoyancy-regimes.tex}

\engterm{The three buoyancy regimes}:

\begin{itemize}[leftmargin=*]
  \item \engterm{Fully submerged, denser than the fluid}: the
    body sinks. Its apparent weight in the fluid is $W_{\text{app
    }} = (\rho_{\text{body}} - \rho_f) g V_{\text{body}}$, a
    positive number, directed downward. The body experiences a
    net downward force and accelerates until balanced by a
    constraint (the ocean floor, a cable).
  \item \engterm{Fully submerged, less dense than the fluid}: the
    body rises. The apparent weight is negative; the body
    accelerates upward until it reaches the free surface and
    enters the floating regime.
  \item \engterm{Floating}: the body's weight equals the buoyancy
    force, with $V_{\text{disp}}$ adjusting to balance:
    \[
    V_{\text{disp}} = \frac{m_{\text{body}}}{\rho_f}.
    \]
    The body floats at a depth such that the displaced volume
    equals the body's mass divided by the fluid density.
\end{itemize}

For a ship in water, $V_{\text{disp}}$ is the volume of water
displaced by the submerged portion of the hull; in standard
naval-architecture units, the ship's \engterm{displacement} is
this mass, typically reported in tonnes or long tons.

\engterm{The stability of a floating body}. A floating body in
static equilibrium can be stable (returning to equilibrium after a
small perturbation), neutrally stable (remaining at the perturbed
position), or unstable (continuing to rotate away from
equilibrium). The stability is set by the geometry of the
displaced volume in the perturbed configuration.

For a body rotated by a small angle $\theta$ from its
equilibrium attitude, the centre of buoyancy shifts. The line of
action of the buoyancy force is vertical and passes through the
shifted centre of buoyancy. The intersection of this line with
the body's centreline (extended through the body) is the
\engterm{metacentre} $M$. The metacentre's position relative to
the body's centre of mass $G$ determines stability:

\begin{itemize}[leftmargin=*]
  \item If $M$ is above $G$ (positive \engterm{metacentric
    height} $\overline{GM} > 0$), the buoyancy force produces a
    restoring moment that returns the body to its equilibrium
    attitude. The body is stable.
  \item If $M$ is below $G$ (negative metacentric height
    $\overline{GM} < 0$), the buoyancy force produces an
    overturning moment that drives the body further from
    equilibrium. The body is unstable.
  \item If $M$ coincides with $G$ ($\overline{GM} = 0$), the
    body is neutrally stable.
\end{itemize}

The metacentric height is computed from the geometry:
\[
\overline{GM} = \overline{BM} - \overline{BG} = \frac{I_{\text{wp
}}}{V_{\text{disp}}} - \overline{BG},
\]
where $I_{\text{wp}}$ is the second moment of the body's
waterplane area (the cross-section of the body at the water
surface) about its longitudinal axis through the waterplane
centroid, $\overline{BM}$ is the distance from the centre of
buoyancy to the metacentre, and $\overline{BG}$ is the distance
from the centre of buoyancy to the centre of mass. Archimedes
analysed the stability of floating paraboloids in Book II of
\emph{On Floating Bodies} \cite[Book II]{
hist:archimedes-floating}, the founding theoretical treatment of
the question; the modern naval-architecture formulation, with
$I_{\text{wp}}$ and $V_{\text{disp}}$, dates to the eighteenth
century.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-metacentre.tex}

\begin{mastery}
\textbf{Where $\overline{BM} = I_{\text{wp}}/V_{\text{disp}}$ comes
from.} Heel the body by a small angle $\theta$. The waterplane (the
cross-section the body makes with the water surface) does not change to
first order, but a wedge of fluid on the rising side leaves the water
and an equal wedge on the descending side enters it. Consider a strip
of the waterplane at transverse distance $y$ from the roll axis, of
area $dA$. The vertical thickness of the wedge there is $y\theta$, so
the volume transferred is $y\theta\,dA$, and the volume entering on one
side equals the volume leaving on the other because $\int y\,dA = 0$
about the centroidal axis. The transferred volume shifts the centre of
buoyancy sideways. The moment of the transferred volume about the roll
axis is
\[
\delta = \int_{\text{waterplane}} (y\theta\,dA)\, y
= \theta \int y^2 \,dA = \theta\, I_{\text{wp}},
\]
where $I_{\text{wp}} = \int y^2\,dA$ is the second moment of the
waterplane area about the roll axis. The horizontal shift of the centre
of buoyancy is this moment divided by the total displaced volume,
$\overline{BB'} = \theta I_{\text{wp}}/V_{\text{disp}}$. For a small
angle the metacentre $M$ lies at $\overline{BB'}/\theta$ above $B$ along
the original vertical, so
\[
\overline{BM} = \frac{\overline{BB'}}{\theta}
= \frac{I_{\text{wp}}}{V_{\text{disp}}}.
\]
The result explains why beam matters so much more than length for roll
stability: $I_{\text{wp}}$ for a rectangular waterplane of length $L$
and beam $B$ is $LB^3/12$, cubic in the beam. Widening a hull by ten
per cent raises $\overline{BM}$ by about a third; lengthening it by ten
per cent raises it only ten per cent. The same wedge argument applied
about the longitudinal axis gives the longitudinal metacentric height,
which is far larger because the ship is long, and is why ships pitch
through small angles but can roll through large ones.
\end{mastery}

The working consequence: a ship with a beamy hull (large
waterplane second moment relative to its displacement) has a
large positive metacentric height and is stable; a tall narrow
hull with the same displacement has a small or negative
metacentric height and may capsize. The stability calculation
is one of the load-bearing analyses of every ship design.

\engterm{Capsize and the angle of vanishing stability}. The
metacentric-height analysis is a small-angle result; at larger
heel angles the geometry of the displaced volume changes
substantially, and the restoring moment must be evaluated by
direct integration. A typical stability diagram (the
\engterm{righting-arm curve} $\overline{GZ}(\theta)$) rises with
angle from a small positive value, reaches a maximum somewhere
between $30^{\circ}$ and $60^{\circ}$ of heel, and then falls,
crossing zero at the \engterm{angle of vanishing stability}.
Beyond this angle the body capsizes. The angle of vanishing
stability is a primary design parameter for vessels expected to
encounter large heel angles in service: lifeboats, fishing
trawlers, small craft in heavy weather.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-righting-arm.tex}

The metacentric height fixes only the slope of the righting-arm curve at
the origin; the full curve, and the angle of vanishing stability that
ends it, depend on the hull shape at heel and on the height of the
deck edge and bilge. Two hulls with the same $\overline{GM}$ can have
very different reserves of stability: a hull with high freeboard keeps
its righting arm rising to large angles, while a hull whose deck edge
immerses early loses righting arm sharply once the edge goes under. The
area under the curve up to the downflooding angle is the energy the hull
can absorb from a gust or a boarding wave before capsize, the quantity a
stability standard actually limits.

\begin{principle}[Floating stability is set by metacentric height]
A floating body is stable against small-angle perturbations if
its metacentric height $\overline{GM} = I_{\text{wp}}/V_{\text{
disp}} - \overline{BG}$ is positive. A wide hull (large
$I_{\text{wp}}$) is stable; a tall hull with mass high in the
hull (large $\overline{BG}$) is unstable. The same geometry
governs the stability of every floating body, from a soap dish
to an aircraft carrier; the working calculation is the same.
\end{principle}

\section{Fluid statics in non-inertial frames}\pathtag{standard}

A fluid in a non-inertial frame (an accelerating tank, a
rotating bucket, a centrifuge) is in static equilibrium in the
frame, with the apparent body force replacing gravity. The
hydrostatic equation generalises directly:
\[
\nabla p = \rho \vec{g}_{\text{eff}}, \quad \vec{g}_{\text{eff}}
= \vec{g} - \vec{a},
\]
with $\vec{a}$ the acceleration of the frame relative to an
inertial frame. The free surface of the fluid is normal to
$\vec{g}_{\text{eff}}$.

\engterm{The linearly accelerating tank}. A rectangular tank
accelerating horizontally at $\vec{a} = a\hat{i}$ has effective
gravity $\vec{g}_{\text{eff}} = -g\hat{k} - a\hat{i}$, tilted from
vertical by an angle $\tan^{-1}(a/g)$. The free surface tilts by
the same angle. A car decelerating at $a = 4\,\text{m/s}^2$
($\tan^{-1}(4/9.81) \approx 22^{\circ}$) produces a surface tilt
of about $22^{\circ}$ in any open container of liquid in the
car; the coffee in an unlidded cup slides toward the front of
the cup.

\engterm{The rotating bucket} (Newton, \emph{Principia}, Book I,
Definition VIII). A cylindrical bucket rotating about its
vertical axis at angular velocity $\omega$ produces a centripetal
acceleration $\omega^2 r$ at radius $r$, directed inward. In the
rotating frame, the apparent body force is $g\hat{k} + \omega^2 r
\hat{r}$ (gravity downward plus centrifugal outward). The free
surface is normal to this apparent gravity at every point and is
therefore a paraboloid:
\[
z(r) = z_0 + \frac{\omega^2 r^2}{2 g}.
\]
The same result governs the figure of a self-gravitating rotating
fluid body (Earth, Jupiter, a rotating drop of mercury); the
paraboloidal-mirror technique of optical-telescope manufacture
(used by Borra and others in the 1990s) exploits the same
hydrostatic shape to produce parabolic mirrors of arbitrarily
high optical quality without grinding.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-rotating-paraboloid.tex}

\engterm{Pressure variation in a rotating fluid}. Integration of
the hydrostatic equation in the rotating frame gives
\[
p(r, z) = p_0 - \rho g z + \tfrac{1}{2} \rho \omega^2 r^2,
\]
with the pressure increasing both with depth (the standard
gravity contribution) and radially outward (the centrifugal
contribution). The result is the working formula for centrifuge
design: a laboratory ultracentrifuge spinning at $\omega \sim
10^4\,\text{rad/s}$ at radius $0.1\,\si{\meter}$ produces an
effective gravity of $\omega^2 r \approx 10^7\,\text{m/s}^2$,
roughly a million $g$, with corresponding pressure gradients
that separate biological molecules by density.

\begin{mastery}
\textbf{Pressure on the wall of an accelerating tank.} Take a closed
rectangular tank of length $\ell$ along the direction of motion, filled
to depth $d$ and accelerating horizontally at $a$. The effective gravity
$\vec{g}_{\text{eff}} = -g\hat{k} - a\hat{i}$ tilts the free surface (or,
in a full closed tank, the surface of constant pressure) by
$\alpha = \tan^{-1}(a/g)$ toward the rear. The pressure field is linear
in both coordinates,
\[
p(x,z) = p_0 - \rho a\,(x - x_0) - \rho g\,(z - z_0),
\]
so the wall pressures differ front to rear. The rear wall carries a
higher hydrostatic load: the surface there is deeper by
$\ell\tan\alpha$. For a road tanker of $\ell = 8\,\si{\meter}$
half-filled with water, braking hard at $a = 6\,\text{m/s}^2$
($\alpha \approx 31^{\circ}$), the surface tips by nearly five metres of
run across the tank, which is why partly filled tankers are placarded and
why baffles are fitted: the longitudinal pressure imbalance, and the
slosh that accompanies a change of acceleration, shift the load and can
unsettle the vehicle. The static analysis gives the steady-state wall
load; the transient slosh is a dynamics problem outside fluid statics,
but the static result already shows that a half-full tank is the worst
case, because a full tank has no free surface to migrate and an empty
one has no mass to migrate.
\end{mastery}

\section{Surface tension and capillarity}\pathtag{mastery}

At a fluid interface (liquid-gas or liquid-liquid), the molecules
at the interface experience an asymmetric attraction from the
two phases. The asymmetry produces a tangential force per unit
length of any line drawn on the interface, the
\engterm{surface tension} $\sigma$. Surface tension has units of
$\text{N}/\text{m}$, equivalently $\text{J}/\text{m}^2$
(energy per unit area, the work required to create new interface
area). Working values: water-air at $20^{\circ}\text{C}$,
$\sigma \approx 0.073\,\text{N}/\text{m}$; mercury-air,
$\sigma \approx 0.485\,\text{N}/\text{m}$; clean steel-air
(solid surface energy, an analogous quantity), $\sigma \approx
2\,\text{N}/\text{m}$.

\engterm{The Young-Laplace equation}. A curved fluid interface
supports a pressure jump between the two sides, of magnitude
\[
\Delta p = \sigma \left(\frac{1}{R_1} + \frac{1}{R_2}\right),
\]
where $R_1$ and $R_2$ are the two principal radii of curvature of
the interface, with sign convention that $R > 0$ when the
interface curves into the higher-pressure side. The equation,
due to Young (1805) and Laplace (1806), is the working formula
for every curved-interface pressure calculation: a spherical
soap bubble of radius $R$ has internal gauge pressure $4\sigma/R$
(two interfaces, each contributing $2\sigma/R$); a spherical
water droplet of radius $R$ has internal gauge pressure
$2\sigma/R$; a cylindrical jet of radius $R$ has internal gauge
pressure $\sigma/R$.

\engterm{Capillary rise}. A wettable tube (a glass capillary in
water) dipped into a fluid sees the fluid rise into the tube to
a height set by the balance between surface tension and gravity:
\[
h = \frac{2\sigma \cos\theta_c}{\rho g r},
\]
with $\theta_c$ the contact angle (zero for perfect wetting,
$90^{\circ}$ for no wetting). The same formula with $\cos\theta_c
< 0$ gives capillary depression for a non-wetting fluid (mercury
in glass). The rise is inversely proportional to the tube radius:
in a $1\,\si{\milli\meter}$ glass capillary, water rises about
$15\,\si{\milli\meter}$; in a $10\,\si{\micro\meter}$ capillary,
water rises about $1.5\,\si{\meter}$.

The capillary-rise formula follows from a force balance that is worth
writing out, because it exposes what the contact angle carries. Around the
circumference of the tube, of length $2\pi r$, the surface tension pulls on
the raised column along the wall at the contact angle $\theta_c$ to the
vertical, so the upward force is $2\pi r\,\sigma\cos\theta_c$. That force
supports the weight of the raised column, $\rho g\,(\pi r^2)\,h$ for a
column of height $h$. Setting the two equal and solving for $h$ returns
$h=2\sigma\cos\theta_c/(\rho g r)$, the result Jurin stated empirically in
1718 before the surface-tension mechanism was understood, which is why the
inverse-radius law still carries his name. The contact angle is a property
of the three phases meeting at the line, solid, liquid, and vapour, and
its cosine sets both the sign and the size of the rise: a perfectly
wetting liquid ($\theta_c=0$) rises fullest, a liquid at $\theta_c=
90^{\circ}$ neither rises nor falls, and a non-wetting liquid
($\theta_c>90^{\circ}$, mercury on glass at about $140^{\circ}$) is
depressed below the outside level. The two behaviours sit side by side in
Figure~\ref{fig:vol03ch10:capillary}: the concave meniscus of a wetting
liquid that climbs the tube, and the convex meniscus of a non-wetting
liquid that is pushed down it.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-capillary.tex}

A working caution the clean-tube formula hides is contact-angle
hysteresis. A real solid surface does not present a single contact angle;
the advancing angle, measured as liquid creeps forward onto dry solid, is
larger than the receding angle, measured as liquid retreats off wetted
solid, and the gap between them can be tens of degrees on a rough or
contaminated surface. The consequence for measurement is that a capillary
rise depends on whether the meniscus arrived by rising or by draining, so a
capillary viscometer or a porometer is read on a single approach direction
and the tube is cleaned to a repeatable wetting state. The same hysteresis
is why a raindrop can sit pinned on an inclined pane, its lower edge at the
advancing angle and its upper edge at the receding angle supplying the
retention force that gravity must overcome before the drop slides. Surface
tension enters fluid statics as a boundary condition on the interface, and
the contact angle is the part of that condition the surface chemistry, not
the bulk fluid, controls.

\engterm{When does surface tension matter?} The dimensionless
ratio
\[
\text{Bo} = \frac{\rho g L^2}{\sigma}
\]
(the \engterm{Bond number}) compares gravity to surface tension
on a length scale $L$. For $\text{Bo} \gg 1$, gravity dominates
and the standard fluid statics applies; for $\text{Bo} \ll 1$,
surface tension dominates and the interface shape is set by the
Young-Laplace equation. The crossover length, where $\text{Bo} =
1$, is the \engterm{capillary length} $\ell_c = \sqrt{\sigma/(
\rho g)}$, which for water-air is about $2.7\,\si{\milli\meter}$.
Surface tension matters for water droplets, capillary tubes,
biological membranes, and microfluidic channels; it does not
matter for ship hulls, dams, or oil tanks. The discipline is to
check the Bond number before deciding whether to include the
surface-tension term.

\section{Pressure in the atmosphere}\pathtag{standard}

The Earth's atmosphere is a compressible fluid in approximately
static equilibrium under gravity (winds and weather systems
introduce small departures that are unimportant for the static
column). The hydrostatic equation in the form $dp/dz = -\rho g$,
combined with the ideal-gas equation $\rho = pM/(RT)$ and a
temperature profile $T(z)$, gives the pressure-altitude relation.

\engterm{The isothermal atmosphere}. With $T(z) = T_0$ constant,
$p(z) = p_0 \exp(-z/H)$, where $H = RT_0/(Mg) \approx 8.4\,
\si{\kilo\meter}$ for the Earth at $T_0 = 288\,\text{K}$. The
isothermal model gives the right scale but is inaccurate over
altitudes where the temperature changes substantially.

\engterm{The standard atmosphere}. The \engterm{International
Standard Atmosphere} (ISA), defined by ICAO Document 7488 and
identical in its first $32\,\si{\kilo\meter}$ to the U.S.\
Standard Atmosphere (1976), specifies a piecewise-linear
temperature profile with a tropospheric lapse rate of $-6.5\,
\si{\kelvin\per\kilo\meter}$ from sea level ($T_0 = 288.15\,
\text{K}$, $p_0 = 101.325\,\text{kPa}$) up to the tropopause at
$11\,\si{\kilo\meter}$, followed by an isothermal stratosphere
($216.65\,\text{K}$) up to $20\,\si{\kilo\meter}$, then a positive
lapse rate to higher altitudes. Within a linear-lapse-rate
layer with constant lapse rate $L = -dT/dz$,
\[
p(z) = p_b \left(\frac{T_b - L(z - z_b)}{T_b}\right)^{Mg/(RL)},
\]
where $p_b, T_b, z_b$ are the values at the base of the layer.
The exponent $Mg/(RL)$ for the tropospheric values is $\approx
5.256$, which gives $p \approx 26.5\,\text{kPa}$ at $10\,\si{
\kilo\meter}$ and $\approx 22.6\,\text{kPa}$ at the tropopause.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-atmosphere.tex}

The layer-by-layer integration is carried out in
\texttt{code/standard\_atmosphere.py}, and its output is tabulated in
\texttt{data/isa\_profile.csv}; the script also computes the
single-scale-height isothermal model for comparison, the contrast that
Figure~\ref{fig:vol03ch10:atmosphere} plots.

The standard atmosphere is the reference against which altimeters,
aircraft pressure-altitude indicators, and engine performance
data are calibrated. The real atmosphere departs from the
standard by amounts that matter in aircraft operations: a
$30\,\text{hPa}$ surface-pressure offset corresponds to a
$\sim 250\,\si{\meter}$ altimeter error at $10\,\si{\kilo\meter}$,
which is why pre-flight altimeter settings (QNH, QFE, QNE) are
part of every flight briefing.

\engterm{Pressure-driven weather signal}. Surface barometric
pressure is a load-bearing input to weather forecasting. Standard
sea-level pressure is approximately $1013\,\text{hPa}$ in
quiescent conditions; high-pressure systems exceed $1030\,\text{
hPa}$, low-pressure systems may fall below $980\,\text{hPa}$, and
tropical cyclones can produce surface pressures below $880\,\text
{hPa}$ at the eye. The pressure record at the eye of Typhoon Tip
(October 1979) of $870\,\text{hPa}$ is the lowest sea-level
pressure ever measured. A barometer at a fixed location,
reporting pressure to $\pm 0.5\,\text{hPa}$ ($\pm 50\,\text{Pa}$),
identifies the passage of weather systems and is the basis of
the project for this chapter.

\engterm{Pressure altitude and density altitude}. An aircraft altimeter is
a barometer with its scale marked in metres against the standard-atmosphere
pressure-height relation, so it reports \engterm{pressure altitude}: the
altitude at which the sensed pressure would occur in the standard
atmosphere, not the true geometric height. Two corrections separate the
reading from the truth. The first is the surface-pressure setting. On a day
when the sea-level pressure differs from the standard $1013\,\text{hPa}$,
the whole pressure-height curve shifts, and an altimeter left on the
standard setting reads high on a low-pressure day, the origin of the
airman's caution ``from high to low, look out below.'' Setting the
subscale to the local sea-level pressure (QNH) makes the altimeter read
height above sea level; setting it to the field pressure (QFE) makes it
read height above the field. The second correction is temperature. The
pressure-height relation the altimeter assumes is built on the standard
temperature profile, and on a day colder than standard the real air is
denser and the true height for a given pressure is lower than indicated,
so cold-weather altimetry carries a further correction that matters for
terrain clearance on an instrument approach. \engterm{Density altitude},
the altitude at which the local air density would occur in the standard
atmosphere, is the quantity that governs engine and wing performance
rather than the instrument reading: on a hot day at a high-elevation field
the density altitude can exceed the geometric elevation by a kilometre or
more, lengthening the takeoff run and cutting the climb rate, which is why
performance charts are entered with density altitude and not with the
number in the altimeter window.

\begin{mastery}
\textbf{Why the atmosphere sorts by molar mass, and why it does not.} The
barometric law $p=p_0\exp(-Mgz/RT)$ carries the molar mass $M$ in the
exponent, so read literally it predicts that each gas in a mixture settles
with its own scale height: the heavy gases (oxygen, argon, carbon dioxide)
would concentrate low and the light gases (helium, hydrogen) would float
high, each following $p_i=p_{i,0}\exp(-M_igz/RT)$ for its own partial
pressure. A still, non-mixing column would indeed sort this way given
enough time, and the scale heights differ substantially: nitrogen's is
about $8.5\,\si{\kilo\meter}$ and hydrogen's roughly fifteen times larger.
The lower atmosphere does not sort, and the reason is that turbulent mixing
overwhelms molecular diffusion below about $100\,\si{\kilo\meter}$. Eddies
stir the whole air parcel, heavy and light molecules together, far faster
than diffusion can separate them, so the composition stays uniform and the
column behaves as a single gas of the mean molar mass $M\approx 0.029\,\si{
\kilogram\per\mole}$. This well-mixed region is the \engterm{homosphere}.
Above it, where the air is too thin for turbulence to keep up, diffusion
wins and the gases do separate by mass into the \engterm{heterosphere},
with light species increasingly dominant at greater height. The engineering
reading is that the single-molar-mass barometric formula is correct in the
homosphere not because gases cannot separate but because mixing prevents
them from doing so, and a design that assumed gravitational sorting of a
gas mixture in an ordinary vessel or room, a stagnant helium layer at the
ceiling, say, would be wrong for the same reason: convection and diffusion
homogenise the mixture long before gravity can stratify it.
\end{mastery}

\begin{estimation}
\textbf{Estimate the hydrostatic force on a typical residential
basement wall, $2.5\,\si{\meter}$ high and $5\,\si{\meter}$
long, with groundwater rising to a depth $1.5\,\si{\meter}$ below
the top of the wall.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
The water column against the wall extends from $1.5\,\si{\meter}$
below the top of the wall ($z_T$ from the top, $z_T = 1.5\,\si{
\meter}$, where the pressure is zero gauge) to the wall's full
depth of $2.5\,\si{\meter}$ from the top (at depth $2.5 - 0 =
2.5\,\si{\meter}$ from grade, equivalent to a water column of
$1\,\si{\meter}$). The wetted area is the wall length times the
water column: $A = 5 \times 1 = 5\,\text{m}^2$. The pressure at
the centroid of the wetted area (at depth $0.5\,\si{\meter}$
below the water table) is $p_C = \rho g h_C = 10^3 \times 9.81
\times 0.5 \approx 5\,\text{kPa}$. The hydrostatic force on the
wetted area is
\[
F = p_C A \approx 5 \times 10^3 \times 5 = 25\,\text{kN}.
\]
The centre of pressure is at depth $h_P = h_C + b^2/(12 h_C)
= 0.5 + 1/(12 \times 0.5) \approx 0.5 + 0.17 \approx 0.67\,\si{
\meter}$ below the water table, or at $1.5 + 0.67 = 2.17\,\si{
\meter}$ from the top of the wall.

That force is significant but not catastrophic: a wall designed
to hold back $25\,\text{kN}$ over a $5\,\si{\meter}$ length is a
standard residential reinforced-concrete basement wall, with
the rebar designed against the bending moment about the wall's
base.

The postmortem on the estimate: the worry case is not the
slow-rise hydrostatic load but the rapid-rise load during heavy
rainfall, when the water table reaches the grade and the entire
$2.5\,\si{\meter}$ height of wall is wetted. The hydrostatic
force then becomes $F = \rho g (b/2) (b \times L) = 10^3 \times
9.81 \times 1.25 \times 12.5 \approx 153\,\text{kN}$ over the
same wall, with the centre of pressure at $2/3$ of the wall
height. The factor of six between dry-basement and fully-wetted
loading is the structural design margin that residential codes
require, and is the reason basement walls in flood-prone areas
are designed for full-saturation loading rather than
typical-water-table loading. The estimate has identified the
worst case as the design-driving case, which is the engineering
discipline the estimation exercise exists to install.
\end{estimation}

\section{Worked examples}\pathtag{standard}

The examples below carry the chapter's working results through complete
calculations, each in the order an engineer would take them: identify
the fluid element or surface, write the balance, integrate, and check
the number against an independent estimate. Each ties to a code file
that reproduces the arithmetic.

\subsection{A mercury barometer and the height of a water barometer}

A simple mercury barometer is a sealed tube, closed at the top and
inverted in a dish of mercury, with vacuum (strictly, mercury vapour at
its low vapour pressure) above the column. The atmospheric pressure on
the dish surface supports the column. At sea level on a standard day,
$p_{\text{atm}} = 101.325\,\text{kPa}$, and the column height is
\[
h = \frac{p_{\text{atm}}}{\rho_{\text{Hg}}\, g}
= \frac{101\,325}{13\,534 \times 9.81}
\approx 0.763\,\si{\meter}.
\]
The familiar $760\,\si{\milli\meter}$ of mercury follows once the
slightly different standard density at $0^{\circ}\text{C}$ is used.
Repeat the calculation with water in place of mercury:
\[
h_{\text{water}} = \frac{101\,325}{998 \times 9.81}
\approx 10.35\,\si{\meter}.
\]
A water barometer would stand over ten metres tall, which is why
Torricelli used mercury: the working fluid's density sets the
instrument's size. The same number, ten metres of water per atmosphere,
is the depth at which a diver has doubled the absolute pressure on the
body, and the maximum height to which a suction pump can lift water (a
pump cannot pull; it can only let the atmosphere push, and the
atmosphere pushes at most ten metres of water column).

\subsection{Force and centre of pressure on a vertical dam}

A straight vertical dam retains water to its crest at height
$H = 30\,\si{\meter}$ over a length $L = 200\,\si{\meter}$. Take the
water surface at the crest, gauge pressure zero. The pressure at depth
$h$ is $p = \rho g h$, rising from zero at the surface to $\rho g H$ at
the base. The resultant per unit width is the area of the triangular
pressure prism,
\[
\frac{F}{L} = \int_0^H \rho g h \,dh = \tfrac12 \rho g H^2
= \tfrac12 \times 998 \times 9.81 \times 30^2
\approx 4.41 \times 10^6\,\text{N/m},
\]
so the total horizontal force is $F = (F/L)\,L \approx 8.81 \times
10^8\,\text{N} \approx 881\,\text{MN}$. The line of action sits at the
centroid of the triangular prism, $H/3 = 10\,\si{\meter}$ above the
base. The overturning moment about the base toe is
\[
M = F \times \frac{H}{3}
\approx 8.81 \times 10^8 \times 10
\approx 8.8 \times 10^9\,\text{N}\cdot\text{m}.
\]
A gravity dam resists this moment with its own weight: the dam must be
massive enough that the weight acting through its base, times its lever
arm to the toe, exceeds the overturning moment with a margin. The
quadratic growth of $F$ with $H$ and the cubic growth of $M$ with $H$
are why dam cross-sections thicken so steeply toward the base. The
numbers here are reproduced by \texttt{code/submerged\_surface.py}.

\subsection{Will it float, and is it stable?}

A homogeneous rectangular box, $L = 10\,\si{\meter}$ long, $B = 4\,\si{
\meter}$ in beam, $D = 2\,\si{\meter}$ deep, built of material at
$\rho_b = 500\,\si{\kilogram\per\meter\cubed}$, is set in fresh water.
The draft follows from vertical equilibrium, weight equals buoyancy:
\[
\rho_b\, (LBD)\, g = \rho_f\, (LBT)\, g
\quad\Rightarrow\quad
T = \frac{\rho_b}{\rho_f} D = \frac{500}{1000} \times 2 = 1.0\,\si{\meter}.
\]
The box floats with one metre submerged and one metre of freeboard. For
stability, compute the metacentric height. The centre of buoyancy is at
half the draft above the keel, $\overline{KB} = T/2 = 0.5\,\si{\meter}$;
the centre of mass of the homogeneous box is at half its height,
$\overline{KG} = D/2 = 1.0\,\si{\meter}$. The waterplane second moment
about the roll (longitudinal) axis is $I_{\text{wp}} = L B^3 / 12 =
10 \times 4^3 / 12 \approx 53.3\,\text{m}^4$, and the displaced volume
is $V_{\text{disp}} = LBT = 40\,\text{m}^3$, so
\[
\overline{BM} = \frac{I_{\text{wp}}}{V_{\text{disp}}}
= \frac{53.3}{40} \approx 1.33\,\si{\meter},
\quad
\overline{GM} = \overline{KB} + \overline{BM} - \overline{KG}
= 0.5 + 1.33 - 1.0 \approx 0.83\,\si{\meter}.
\]
The metacentric height is positive, so the box is stable in roll. Repeat
the calculation for a tall narrow box, $L = 10$, $B = 1$, $D = 4$,
same density: the draft is $T = 2\,\si{\meter}$, $\overline{KB} = 1.0$,
$\overline{KG} = 2.0$, $I_{\text{wp}} = 10 \times 1^3/12 \approx 0.83\,
\text{m}^4$, $V_{\text{disp}} = 20\,\text{m}^3$, so $\overline{BM}
\approx 0.042\,\si{\meter}$ and $\overline{GM} \approx 1.0 + 0.042 -
2.0 \approx -0.96\,\si{\meter}$, negative. The tall box capsizes the
instant it is released, even though it floats. Beam, entering the
stability through $B^3$, is the variable that buys roll stability. The
two cases and several others are tabulated in
\texttt{data/box\_stability.csv} and reproduced by
\texttt{code/metacentre.py}.

\subsection{Reading a differential manometer across an orifice}

A horizontal water pipe carries a mercury differential manometer across
an orifice plate; the mercury stands $\Delta h = 50\,\si{\milli\meter}$
higher in the downstream leg. The connecting lines are full of the
process water. Tracing the hydrostatic path from the high-pressure tap
down through water, across the mercury, and up through water to the
low-pressure tap, the equal water columns cancel and the working
equation is
\[
p_1 - p_2 = (\rho_{\text{Hg}} - \rho_{\text{water}})\, g\, \Delta h
= (13\,534 - 998) \times 9.81 \times 0.050
\approx 6.15\,\text{kPa}.
\]
The process-fluid density enters with a minus sign because the water
fills the lines above the mercury on both legs; dropping that term (a
common slip) would overstate the differential by about eight per cent.
The reduction is checked in \texttt{code/manometer.py}.

\subsection{Force and centre of pressure on a curved gate}

A radial sluice gate (a Tainter gate) is a curved plate shaped as part
of a cylinder, hinged at the centre of its arc, that holds back water in
a spillway. Take a gate that is a quarter-circle of radius $R = 4\,\si{
\meter}$ and width $w = 6\,\si{\meter}$, with the water surface at the
top of the arc and the arc sweeping from the surface down to a point
$R$ directly below the hinge. We want the horizontal and vertical
components of the hydrostatic force and the resultant.

The horizontal component acts on the vertical projection of the curved
surface, which is a rectangle of height $R$ and width $w$ with its top
at the free surface. Its centroid sits at depth $R/2$, so
\[
F_H = \rho g (R/2)(R w)
= 998 \times 9.81 \times 2 \times (4 \times 6)
\approx 4.70 \times 10^5\,\text{N}
\approx 470\,\text{kN},
\]
acting at the centre of pressure of the projection, $h_P = (R/2) +
(wR^3/12)/[(R/2)(Rw)] = R/2 + R/6 = 2R/3 \approx 2.67\,\si{\meter}$ below
the surface. The vertical component equals the weight of the water that
would fill the volume between the curved surface and the free surface.
That volume is the width times the plane area between the quarter-arc and
its bounding rectangle, which is the quarter-disc area $\tfrac14\pi R^2$:
\[
F_V = \rho g\, w\, \tfrac14 \pi R^2
= 998 \times 9.81 \times 6 \times \tfrac14 \pi \times 16
\approx 7.38 \times 10^5\,\text{N}
\approx 738\,\text{kN},
\]
directed downward on the gate. The resultant magnitude is
\[
F = \sqrt{F_H^2 + F_V^2}
= \sqrt{470^2 + 738^2}\;\text{kN}
\approx 875\,\text{kN},
\]
inclined below the horizontal at $\arctan(F_V/F_H) = \arctan(738/470)
\approx 57.5^{\circ}$. Because every pressure on a circular arc points
along a radius toward the hinge, the resultant line of action passes
through the hinge, and the gate carries no net moment about its own
pivot. That is the design intent of a radial gate: the water load is
reacted as a thrust through the bearing, and the hoist need only lift the
gate's own weight, not fight the water moment. The decomposition is
plotted in Figure~\ref{fig:vol03ch10:curved-surface}.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-curved-surface.tex}

\subsection{A two-fluid manometer with an air gap}

A closed tank holds oil ($\rho_{\text{oil}} = 850\,\si{\kilogram\per
\meter\cubed}$) with a trapped air space above it. A U-tube manometer
taps the tank below the oil surface and runs to atmosphere through
mercury ($\rho_{\text{Hg}} = 13\,534\,\si{\kilogram\per\meter\cubed}$).
Reading down from the tank tap: $0.40\,\si{\meter}$ of oil, then the
mercury, which stands $0.25\,\si{\meter}$ higher in the open leg than in
the tank leg. We want the gauge pressure of the air space, given that the
oil surface is $0.60\,\si{\meter}$ above the tank tap.

The discipline is to trace one continuous fluid path from the air space
to the atmosphere, adding $\rho g \Delta z$ on every descent and
subtracting it on every rise. Starting at the air space (pressure
$p_{\text{air}}$, the unknown) and descending through $0.60\,\si{\meter}$
of oil to the tank tap, then a further $0.40\,\si{\meter}$ of oil to the
mercury surface in the tank leg, then up $0.25\,\si{\meter}$ of mercury
to the open leg surface at atmosphere:
\[
p_{\text{air}}
+ \rho_{\text{oil}} g (0.60 + 0.40)
- \rho_{\text{Hg}} g (0.25)
= p_{\text{atm}}.
\]
Solving for the gauge pressure $p_{\text{air}} - p_{\text{atm}}$:
\[
p_{\text{air,gauge}}
= \rho_{\text{Hg}} g (0.25) - \rho_{\text{oil}} g (1.00)
= (13\,534 \times 0.25 - 850 \times 1.00) \times 9.81
\approx 24.9\,\text{kPa}.
\]
The mercury column carries the bulk of the reading; the oil column
subtracts about $8.3\,\text{kPa}$, a one-third correction that a single
fluid manometer would miss. Two lessons sit in the arithmetic: the path
trace handles any number of stacked fluids by bookkeeping alone, and the
lighter fluid's column is not negligible when it is as deep as the
denser fluid's height difference. The same path-trace logic runs in
\texttt{code/manometer.py}.

\subsection{Buoyant force on a partially submerged crane load}

A solid steel cylinder, diameter $0.50\,\si{\meter}$ and length $2.0\,
\si{\meter}$ ($\rho_{\text{steel}} = 7850\,\si{\kilogram\per\meter
\cubed}$), is lowered vertically into seawater ($\rho_f = 1025\,\si{
\kilogram\per\meter\cubed}$) by a crane. We want the cable tension as a
function of how far the cylinder has entered the water, and the tension
when it is fully submerged.

The cylinder's full volume is $V = \pi (0.25)^2 (2.0) \approx 0.3927\,
\text{m}^3$ and its weight in air is $W = \rho_{\text{steel}} g V =
7850 \times 9.81 \times 0.3927 \approx 30.2\,\text{kN}$. When a length
$\ell$ has entered the water (measured up from the bottom face), the
submerged volume is $V_{\text{sub}} = \pi (0.25)^2 \ell$ and the buoyant
force is $F_B = \rho_f g V_{\text{sub}} = 1025 \times 9.81 \times \pi
(0.0625) \ell \approx 1.974\,\ell\;\text{kN}$ (with $\ell$ in metres).
The cable tension is the weight minus the buoyancy,
\[
T(\ell) = W - F_B = 30.2 - 1.974\,\ell \quad (\text{kN},\; 0 \le \ell
\le 2.0).
\]
At full submersion, $\ell = 2.0\,\si{\meter}$:
\[
T = 30.2 - 1.974 \times 2.0 \approx 26.3\,\text{kN}.
\]
The cable sheds about $3.9\,\text{kN}$, roughly thirteen per cent of the
dry weight, as the cylinder enters the water; the apparent-weight relief
is the buoyancy of the displaced seawater. The tension falls linearly
with depth of entry because the cylinder has constant cross-section, so
the rigger feels a steady lightening rather than a sudden drop. A crane
plan for a subsea lift uses exactly this curve, because the lightest
tension (full submersion) and the snap loads at the splash zone, where a
wave can momentarily expose and re-submerge the load, bracket the cable
rating.

\subsection{Force on a quarter-circle gate by components, water on the underside}

The Tainter-gate example took a curved surface with water pressing on its
convex face, where the vertical component is the weight of real water
standing above the gate. The complementary case puts the water on the
concave underside, where the vertical component is the weight of the
\emph{imaginary} water the surface would have to displace, and it points
upward. A quarter-circle gate of radius $R = 3\,\si{\meter}$ and width
$w = 5\,\si{\meter}$ closes the bottom of a tank; the curved surface runs
from a vertical tangent at the tank floor up to a horizontal tangent at the
free surface, and the water fills the region inside the arc, pressing
outward and upward on the gate. The free surface is at the top of the arc.

The horizontal component acts on the vertical projection of the curved
surface, a rectangle of height $R$ and width $w$ with its top at the free
surface, centroid at depth $R/2$:
\[
F_H = \rho g (R/2)(R w)
= 1000 \times 9.81 \times 1.5 \times (3 \times 5)
\approx 2.21 \times 10^5\,\text{N}
\approx 221\,\text{kN}.
\]
The vertical component equals the weight of water in the volume bounded by
the curved surface, the free surface, and the verticals, which for a
concave-up gate is the quarter-disc footprint that the real water occupies
above the gate, $\tfrac14 \pi R^2$ in section, and the force points
\emph{upward} on the gate:
\[
F_V = \rho g\, w\, \tfrac14 \pi R^2
= 1000 \times 9.81 \times 5 \times \tfrac14 \pi \times 9
\approx 3.47 \times 10^5\,\text{N}
\approx 347\,\text{kN},
\]
directed up. The resultant magnitude is
\[
F = \sqrt{F_H^2 + F_V^2}
= \sqrt{221^2 + 347^2}\;\text{kN}
\approx 411\,\text{kN},
\]
inclined above the horizontal at $\arctan(F_V/F_H) = \arctan(347/221)
\approx 57.5^{\circ}$. The bookkeeping rule that prevents the common sign
error: the vertical component is the weight of the water column between the
surface and the curved face whether that water is real (pressing down on a
convex face) or imaginary (the volume the concave face would enclose,
giving an upward push). Drawing the column and asking whether real water
sits in it fixes the direction every time.

\subsection{Pressure in a uniformly accelerating fuel tank}

A road tanker carries diesel ($\rho = 840\,\si{\kilogram\per\meter\cubed}$)
in a closed rectangular tank, length $\ell = 7\,\si{\meter}$, height
$2\,\si{\meter}$, filled to a depth of $1.2\,\si{\meter}$ at rest. The
tanker accelerates steadily at $a = 2.5\,\text{m/s}^2$. We want the surface
tilt and the gauge pressure at the rear-bottom corner.

In the tank frame the effective gravity is $\vec g_{\text{eff}} = -g\hat k
- a\hat i$, tilted from vertical toward the rear by
\[
\alpha = \arctan\!\left(\frac{a}{g}\right)
= \arctan\!\left(\frac{2.5}{9.81}\right)
\approx 14.3^{\circ},
\]
so the free surface tips up toward the rear by the same angle, the geometry
of Figure~\ref{fig:vol03ch10:accel-tank}. Over the $7\,\si{\meter}$ length
the surface rises at the rear relative to the front by
\[
\Delta h = \ell \tan\alpha = 7 \times \tan 14.3^{\circ}
\approx 7 \times 0.255 \approx 1.78\,\si{\meter},
\]
which is more than the at-rest free-surface clearance, so part of the front
roof would be exposed and the rear surface would reach the roof: this tank
sloshes hard under that acceleration unless baffled, and a half-fill is the
worst case, exactly the warning of the earlier mastery box. Taking the
tank as full enough that the rear corner stays wetted, the pressure field
is linear in both coordinates,
\[
p(x, z) = p_0 - \rho a (x - x_0) - \rho g (z - z_0).
\]
Reference the surface point at the rear top of the liquid as the zero-gauge
origin and descend to the rear-bottom corner, a vertical drop equal to the
local liquid depth there. With the surface tilted, the rear-corner depth
below its own surface point is the at-rest depth plus the rear-end rise,
about $1.2 + \tfrac12(1.78) \approx 2.0\,\si{\meter}$ measured along the
true vertical for a tank that just contains the tilted surface. The
pressure increases along $\vec g_{\text{eff}}$, whose magnitude is
\[
|\vec g_{\text{eff}}| = \sqrt{g^2 + a^2}
= \sqrt{9.81^2 + 2.5^2}
\approx 10.1\,\text{m/s}^2,
\]
so the gauge pressure at the rear-bottom corner, taken as the deepest point
measured along the effective-gravity direction over a slant depth of about
$2.0\,\si{\meter}$, is
\[
p_{\text{rear}} \approx \rho\,|\vec g_{\text{eff}}|\, d_{\text{slant}}
\approx 840 \times 10.1 \times 2.0
\approx 17.0\,\text{kPa}.
\]
The rear wall carries the higher load; the front wall, where the surface
has dropped, carries less. The asymmetry is why a partly filled tanker
behaves differently under braking and acceleration, and why baffles and
load placarding exist: the static result alone shows the rear-heavy
pressure shift, before any transient slosh is considered.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-accel-tank.tex}

\subsection{A floating hydrometer and the stability of a floating stick}

A hydrometer measures liquid density by the depth to which a weighted stem
floats: the denser the liquid, the higher the instrument rides. Take a
hydrometer with a bulb of volume $V_b = 4.0\,\text{cm}^3$ and a thin
cylindrical stem of cross-section $A_s = 0.20\,\text{cm}^2$, total mass
$m = 4.2\,\si{\gram}$. In water ($\rho_w = 1000\,\si{\kilogram\per\meter
\cubed}$) the instrument floats with the bulb fully submerged and the stem
partly out; the submerged volume that balances the weight is
\[
V_{\text{disp}} = \frac{m}{\rho_w}
= \frac{4.2 \times 10^{-3}}{1000}
= 4.2 \times 10^{-6}\,\text{m}^3
= 4.2\,\text{cm}^3.
\]
The bulb supplies $4.0\,\text{cm}^3$, so the stem is submerged by the
remaining $0.2\,\text{cm}^3$, a stem length of
$\ell_w = 0.2/0.20 = 1.0\,\si{\centi\meter}$ below the waterline. In a
denser brine, $\rho = 1050\,\si{\kilogram\per\meter\cubed}$, the displaced
volume needed falls to $V_{\text{disp}} = 4.2\times10^{-3}/1050
\approx 4.0\,\text{cm}^3$, exactly the bulb, so the waterline sits right at
the bulb-stem junction and the stem rides $1.0\,\si{\centi\meter}$ higher
than in water. The stem displacement per unit density change,
$d\ell/d\rho = m/(\rho^2 A_s)$, is the instrument's sensitivity; a thinner
stem reads finer, the reason precision hydrometers carry hair-thin stems.

The instrument floats stem-up rather than tipping over because of the same
metacentric logic that governs a ship, applied to a slender float. The
mass concentrated low in the bulb puts the centre of gravity well below the
centre of buoyancy, so $\overline{BG}$ is negative in the keel-up sense and
the metacentric height $\overline{GM} = \overline{BM} - \overline{BG}$ is
strongly positive even though the waterplane (the stem cross-section) is
tiny and $\overline{BM} = I_{\text{wp}}/V_{\text{disp}}$ is small. A
floating stick with mass spread uniformly along its length, by contrast,
floats on its side: with the centre of gravity at the geometric centre,
$\overline{BG}$ is positive and the slender waterplane cannot supply enough
$\overline{BM}$ to lift the metacentre above the centre of gravity, so the
upright attitude is unstable and the stick rolls to lie flat, where the
long waterplane gives a large $I_{\text{wp}}$ and a stable float. Ballast
low in the bulb is what lets the hydrometer stand up where the bare stick
lies down.

\subsection{Pressure profile in a three-fluid layered tank}

A storage tank settles into three immiscible layers: a $1.2\,\si{\meter}$
oil cap ($\rho_{\text{oil}} = 850\,\si{\kilogram\per\meter\cubed}$) over a
$2.0\,\si{\meter}$ water layer ($\rho_w = 1000\,\si{\kilogram\per\meter
\cubed}$) over a $1.0\,\si{\meter}$ brine sediment ($\rho_b = 1180\,\si{
\kilogram\per\meter\cubed}$), free surface open to atmosphere. We want the
gauge pressure at each interface and at the floor, and the resultant
horizontal force on the tank wall over the full depth.

The pressure builds layer by layer, each layer adding $\rho g h$ at the
local specific weight, the piecewise-linear profile of
Figure~\ref{fig:vol03ch10:layered-tank}.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-layered-tank.tex}

At the oil-water interface,
\[
p_1 = \rho_{\text{oil}} g\, h_{\text{oil}}
= 850 \times 9.81 \times 1.2 \approx 10.0\,\text{kPa}.
\]
At the water-brine interface, add the water column ($\rho_w g h_w = 1000
\times 9.81 \times 2.0 \approx 19.6\,\text{kPa}$),
\[
p_2 = p_1 + \rho_w g\, h_w
\approx 10.0 + 19.6 \approx 29.6\,\text{kPa}.
\]
At the floor, add the brine column ($\rho_b g h_b = 1180 \times 9.81 \times
1.0 \approx 11.6\,\text{kPa}$),
\[
p_3 = p_2 + \rho_b g\, h_b
\approx 29.6 + 11.6 \approx 41.2\,\text{kPa}.
\]
A single-fluid estimate that took the whole $4.2\,\si{\meter}$ as water
would give $\rho_w g (4.2) \approx 41.2\,\text{kPa}$ at the floor by
coincidence here (the light oil cap and the heavy brine nearly offset), but
the interface pressures would be wrong, and the wall force, which weights
the deeper high-pressure region, would not. The horizontal force per unit
wall width is the area under the pressure profile, summed segment by
segment as the trapezoids of the figure,
\[
\frac{F}{w} = \tfrac12 p_1 h_{\text{oil}}
+ \tfrac12 (p_1 + p_2) h_w
+ \tfrac12 (p_2 + p_3) h_b,
\]
\[
\frac{F}{w} = \tfrac12 (10.0)(1.2) + \tfrac12 (10.0 + 29.6)(2.0)
+ \tfrac12 (29.6 + 41.2)(1.0)
\approx 6.0 + 39.6 + 35.4 \approx 81.0\,\text{kN/m}.
\]
The lesson the layered tank teaches is that pressure is a running sum down
the column and never resets at an interface; the only thing that changes at
each interface is the slope of the profile, which steepens as the denser
fluid takes over. Stacking the columns in the wrong order, or treating the
stack as a single average fluid, gets the floor pressure roughly right by
luck and the interface loads wrong by design.

\subsection{Force on a sloped trapezoidal gate}

Not every gate is a tidy rectangle. A trash sluice on an irrigation canal
is a flat plate shaped as a symmetric trapezoid, wider at the bottom than
at the top, set on a slope. Take a trapezoid with top width $a = 2\,\si{
\meter}$, bottom width $c = 4\,\si{\meter}$, and slant height $b = 3\,\si{
\meter}$ along the plate, inclined at $\theta = 50^{\circ}$ from horizontal,
with its top edge at the free surface. We want the resultant hydrostatic
force and the depth of its centre of pressure.

The plate area is the trapezoid area,
\[
A = \tfrac12 (a + c)\, b = \tfrac12 (2 + 4)(3) = 9\,\text{m}^2.
\]
The centroid of a trapezoid lies, measured along the slant height from the
top (the narrow) edge, at
\[
\bar{s} = \frac{b}{3}\cdot\frac{a + 2c}{a + c}
= \frac{3}{3}\cdot\frac{2 + 8}{6}
= \frac{10}{6} \approx 1.667\,\si{\meter},
\]
so the centroid is below the surface (which is at the top edge) by the
vertical drop
\[
h_C = \bar{s}\,\sin\theta = 1.667 \times \sin 50^{\circ}
\approx 1.667 \times 0.766 \approx 1.28\,\si{\meter}.
\]
The resultant force is the pressure at the centroid times the area,
\[
F = \rho g h_C A
= 1000 \times 9.81 \times 1.28 \times 9
\approx 1.13 \times 10^5\,\text{N}
\approx 113\,\text{kN},
\]
acting normal to the plate. For the centre of pressure we need the second
moment of the trapezoid about its own centroidal axis (parallel to the
parallel sides). A standard result for a trapezoid of parallel sides $a$,
$c$ and height $b$ is
\[
I_C = \frac{b^3\,(a^2 + 4ac + c^2)}{36\,(a + c)}
= \frac{27\,(4 + 32 + 16)}{36 \times 6}
= \frac{27 \times 52}{216}
= 6.5\,\text{m}^4.
\]
The centre of pressure lies below the centroid, measured along the slant,
by $I_C \sin\theta/(h_C A)$, so its depth below the surface is
\[
h_P = h_C + \frac{I_C \sin^2\theta}{h_C A}
= 1.28 + \frac{6.5 \times 0.766^2}{1.28 \times 9}
\approx 1.28 + 0.331 \approx 1.61\,\si{\meter}.
\]
The trapezoid carries its load lower than a rectangle of the same area
would, because the wide bottom places more area in the deeper, higher-
pressure water. The two facts an irregular gate demands, the centroid
location and the second moment of the actual shape, are the only inputs
that change; the $p_C A$ and centre-of-pressure machinery is unchanged.

\subsection{A hydraulic jack and the price of force multiplication}

A bottle jack multiplies a hand force into a lifting force by Pascal's
principle: the same pressure acts on a small pumping piston and a large
ram. Take a jack with a pump-piston diameter $d_1 = 12\,\si{\milli\meter}$
and a ram diameter $d_2 = 40\,\si{\milli\meter}$, with a hand lever of
mechanical advantage $10{:}1$ (the lever multiplies the operator's hand
force tenfold onto the pump piston). The operator pushes the lever end with
$150\,\text{N}$. We want the load the jack can lift and the lever travel
needed to raise the load one centimetre.

The hand force on the pump piston is $F_1 = 10 \times 150 = 1500\,\text{N}$.
The pump piston produces a fluid gauge pressure
\[
p = \frac{F_1}{A_1}
= \frac{1500}{\tfrac{\pi}{4}(0.012)^2}
= \frac{1500}{1.131 \times 10^{-4}}
\approx 1.33 \times 10^7\,\text{Pa}
\approx 13.3\,\text{MPa}.
\]
That pressure acts undiminished on the ram, of area $A_2 = \tfrac{\pi}{4}
(0.040)^2 \approx 1.257 \times 10^{-3}\,\text{m}^2$, lifting
\[
F_2 = p\,A_2 = 1.33 \times 10^7 \times 1.257 \times 10^{-3}
\approx 1.67 \times 10^4\,\text{N}
\approx 16.7\,\text{kN},
\]
a load of about $1.7\,\si{\tonne}$ from a $150\,\text{N}$ hand push, an
overall force multiplication of $F_2/150 \approx 111$. The multiplication
is the lever ratio times the piston-area ratio, $10 \times (40/12)^2
\approx 10 \times 11.1 \approx 111$, and the two factors simply multiply.
The price is travel, and it is paid in full. The incompressible fluid
conserves volume, so the ram rises by the pumped volume divided by the ram
area; to raise the ram $\Delta_2 = 1\,\si{\centi\meter}$ the pump must
displace $A_2 \Delta_2$, which takes a pump-piston stroke of
\[
\Delta_1 = \frac{A_2}{A_1}\Delta_2 = \left(\frac{40}{12}\right)^2
\times 1 \approx 11.1\,\si{\centi\meter},
\]
and the lever end, ten times again, must sweep $10 \times 11.1 \approx
111\,\si{\centi\meter}$ of travel for that one centimetre of lift. Force out
times distance out equals force in times distance in: the hydraulic jack
buys force with distance at a fixed exchange rate, and the work is
conserved exactly because the fluid neither stores nor returns energy in
the static idealisation. The many short lever strokes of a real bottle jack
are the distance side of the same bargain, metered out by a check valve so
the ram holds its gain between strokes.

\subsection{A weather balloon and its ceiling}

A latex sounding balloon is filled with helium at the surface and released.
Unlike a fixed-volume airship, a free latex balloon expands as it rises,
because the latex offers almost no resistance and the helium stays near
ambient pressure; the balloon climbs until it bursts at its design
diameter. We want the net lift at launch and a sense of why a free balloon
keeps rising while a fixed-volume craft levels off, the contrast plotted in
Figure~\ref{fig:vol03ch10:airship-equilibrium}.

At launch the balloon holds $V_0 = 3.0\,\text{m}^3$ of helium. Sea-level
air density is $\rho_a = 1.225\,\si{\kilogram\per\meter\cubed}$ and helium
density is $\rho_{\text{He}} = 0.169\,\si{\kilogram\per\meter\cubed}$. The
gross buoyant force is the weight of displaced air,
\[
F_B = \rho_a g V_0 = 1.225 \times 9.81 \times 3.0 \approx 36.1\,\text{N}.
\]
The helium itself weighs $\rho_{\text{He}} g V_0 = 0.169 \times 9.81
\times 3.0 \approx 5.0\,\text{N}$, so the free lift available to carry
envelope, instrument package, and any payload is
\[
L_{\text{net}} = (\rho_a - \rho_{\text{He}}) g V_0
\approx (1.225 - 0.169) \times 9.81 \times 3.0 \approx 31.1\,\text{N},
\]
enough to lift about $3.2\,\si{\kilogram}$. If the balloon plus payload
masses $1.5\,\si{\kilogram}$ (weight $14.7\,\text{N}$), it launches with a
comfortable surplus and accelerates upward.

Here the free balloon and the sealed airship part ways. A sealed
fixed-volume craft displaces a constant $V$, so as air density falls with
altitude its buoyant force $\rho_a(z) g V$ falls with it and the craft
settles at the pressure ceiling where buoyancy meets weight, the zero
crossing of Figure~\ref{fig:vol03ch10:airship-equilibrium}. A free latex
balloon instead expands to hold ambient pressure, so its volume grows as
the air thins; to first order the displaced mass $\rho_a(z) V(z)$ stays
near constant because $V$ rises as $\rho_a$ falls, and the net lift barely
sags with altitude. The balloon therefore keeps climbing until the latex
reaches its bursting diameter, often near $30\,\si{\kilo\meter}$, where the
envelope tears and the package parachutes back. The design variable for a
fixed-volume airship is the ceiling altitude; for a free sounding balloon
it is the burst diameter, and the same buoyancy balance, read with the
volume held constant or let free, gives the two different behaviours.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-airship-equilibrium.tex}

\subsection{A siphon and the height of its crest}

A siphon carries liquid from a supply reservoir up over a crest and down to
a lower delivery point, running with no pump once it is primed. Take a
siphon of clear-bore tube moving water from a tank whose surface is
$\Delta z = 1.5\,\si{\meter}$ above the delivery surface, with the crest of
the tube $h_c = 6\,\si{\meter}$ above the supply surface, the geometry of
Figure~\ref{fig:vol03ch10:siphon}. We want the maximum crest height the
siphon can tolerate before the column breaks, and where this one stands
against that limit.

The static reasoning ignores the flow speed and asks only whether the
pressure stays positive everywhere along the tube. Pressure is highest at
the two reservoir surfaces (atmospheric) and falls as the liquid is lifted;
the lowest pressure is at the crest, the highest point the liquid reaches.
Tracing the hydrostatic path from the supply surface up to the crest, a
rise of $h_c$, the gauge pressure at the crest is
\[
p_{\text{crest}} = p_{\text{atm}} - \rho g h_c
= 101\,325 - 1000 \times 9.81 \times 6
\approx 101\,325 - 58\,900 \approx 42.4\,\text{kPa absolute}.
\]
The crest pressure is well above zero, so this siphon runs. The ceiling on
crest height is set by the point where the crest pressure would fall to the
liquid's vapour pressure, at which the water boils and a vapour pocket
breaks the column:
\[
h_{c,\max} = \frac{p_{\text{atm}} - p_{\text{vap}}}{\rho g}
= \frac{101\,325 - 2\,340}{1000 \times 9.81}
\approx 10.1\,\si{\meter},
\]
taking the vapour pressure of water at $20^{\circ}\text{C}$ as
$2.34\,\text{kPa}$. The practical ceiling is a little lower still, because
the moving liquid also spends pressure on the velocity head and on
friction, and because dissolved air comes out of solution before the
vapour pressure is reached. Two lessons sit in the arithmetic. The crest
height, not the total drop, sets whether a siphon can run, so a siphon over
a high wall fails even when its delivery point is far below its supply. And
the same ten metres of water that limits a suction pump, a water barometer,
and a siphon crest is one number, $p_{\text{atm}}/\rho g$, wearing three
costumes: in every case the atmosphere can push a water column up by at
most one atmosphere's worth of head.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-siphon.tex}

\subsection{A diving bell and the compression of its air pocket}

An open-bottom diving bell traps a fixed mass of air and carries it down
for a worker to breathe. Take a cylindrical bell of internal
cross-sectional area $A = 3\,\text{m}^2$ and internal height $H = 2.5\,\si{
\meter}$, lowered so its rim reaches a depth of $d = 20\,\si{\meter}$ in
seawater ($\rho_f = 1025\,\si{\kilogram\per\meter\cubed}$). At the surface
the bell is full of air at atmospheric pressure. We want the height of the
air space remaining once the bell settles at depth, the geometry of
Figure~\ref{fig:vol03ch10:diving-bell}.

The trapped air obeys Boyle's law at the roughly constant temperature of
the descent, $p_1 V_1 = p_2 V_2$, so the working step is to find the
pressure of the air at depth. The air-water interface inside the bell sits
somewhere between the rim and the roof; call the air-space height $y$, so
the interface is at depth $d - (H - y)$ below the surface, and the air
pressure equals the water pressure at that interface,
\[
p_2 = p_{\text{atm}} + \rho_f g\,[\,d - (H - y)\,].
\]
With the air volume $V_2 = A y$ and $V_1 = A H$ at the surface, Boyle's law
gives one equation in $y$,
\[
p_{\text{atm}} H = \big(p_{\text{atm}} + \rho_f g\,[d - H + y]\big)\,y.
\]
A first pass ignores the small $(H-y)$ term, taking the air pressure as the
pressure at the rim, $p_2 \approx p_{\text{atm}} + \rho_f g d = 101\,325 +
1025 \times 9.81 \times 20 \approx 302\,\text{kPa}$, about three
atmospheres. Then
\[
y \approx H\,\frac{p_{\text{atm}}}{p_2}
= 2.5 \times \frac{101\,325}{302\,000}
\approx 0.84\,\si{\meter}.
\]
The air compresses to about a third of its surface volume, so water rises
about $1.66\,\si{\meter}$ into the bell and only the top $0.84\,\si{\meter}$
stays breathable. Solving the full quadratic in $y$, which accounts for the
interface sitting above the rim, shifts the answer only slightly, to about
$0.86\,\si{\meter}$, because the $(H-y)$ correction is small against the
$20\,\si{\meter}$ depth. The design response is to pump fresh compressed air
down to blow the water back out and restore the working height, which is
exactly what a pneumatic caisson does to keep a dry working chamber open
under a river while its foundation is dug: the caisson is a diving bell
scaled up to hold a work crew, and its air pressure is set to balance the
water pressure at the digging face.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-diving-bell.tex}

\subsection{An iceberg, its hidden fraction, and the melting-ice question}

A body of ice floats in seawater with a submerged fraction fixed by the
density ratio, the balance of Figure~\ref{fig:vol03ch10:iceberg}. Glacial
ice has density $\rho_i \approx 917\,\si{\kilogram\per\meter\cubed}$ and
seawater $\rho_f \approx 1025\,\si{\kilogram\per\meter\cubed}$, so vertical
equilibrium (weight equals buoyancy) gives the submerged fraction directly,
\[
\frac{V_{\text{sub}}}{V} = \frac{\rho_i}{\rho_f}
= \frac{917}{1025} \approx 0.895.
\]
About ninety per cent of an iceberg rides below the waterline and a tenth
shows, the tip of the iceberg made quantitative. In fresh water the ratio
is $917/1000 \approx 0.917$, so an ice cube in a glass hides even more of
itself.

A cleaner test of the reasoning is the classic question: when the floating
ice melts, does the water level rise, fall, or hold? Take a floating chunk
of mass $m$. While it floats it displaces a volume $V_{\text{disp}} = m/
\rho_f$ of the surrounding water, by Archimedes. When it melts it turns into
a mass $m$ of water, which occupies a volume $m/\rho_f$ at the density of
the surrounding water. The melt water exactly fills the volume the ice had
been displacing, so the level does not change. The result holds for pure
ice in the same water it melts into, and it is why melting sea ice does not
by itself raise sea level. Two riders sharpen it. Ice that carries embedded
rock or sand displaces extra water while floating (the dense inclusion adds
to the displacement) but drops its load to the bottom on melting, so a
dirty berg lowers the level slightly as it melts. And ice resting on land,
a glacier or an ice sheet, displaces no sea water at all until it melts and
runs off, so land ice raises the level while floating ice does not, the
distinction that separates the two contributions to sea-level change.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-iceberg.tex}

\subsection{Force and centre of pressure on a submerged circular gate}

A circular gate is the shape a round pipe or culvert presents to a head of
water, and its second moment differs from the rectangle worked earlier.
Take a vertical circular gate of radius $a = 0.6\,\si{\meter}$ with its
centre at depth $h_C = 4\,\si{\meter}$ below the free surface in fresh
water. We want the resultant force and the depth of the centre of pressure.

The centroid of a circle is its centre, so the centroidal depth is the
given $h_C = 4\,\si{\meter}$ and the area is $A = \pi a^2 = \pi (0.6)^2
\approx 1.131\,\text{m}^2$. The resultant force is the pressure at the
centroid times the area,
\[
F = \rho g h_C A
= 1000 \times 9.81 \times 4 \times 1.131
\approx 4.44 \times 10^4\,\text{N}
\approx 44.4\,\text{kN}.
\]
The second moment of a circle about a horizontal centroidal axis is
$I_{xx,C} = \pi a^4/4 = \pi (0.6)^4/4 \approx 0.1018\,\text{m}^4$, so the
centre of pressure sits below the centroid by
\[
h_P - h_C = \frac{I_{xx,C}}{h_C A}
= \frac{0.1018}{4 \times 1.131}
\approx 0.0225\,\si{\meter},
\]
placing it at $h_P \approx 4.02\,\si{\meter}$. The offset is small, a little
over two centimetres, because the gate is deep relative to its size and the
pressure is nearly uniform across it. The general rule the number
illustrates: the centre-of-pressure offset $I_{xx,C}/(h_C A)$ scales as the
square of the gate size divided by the centroidal depth, so a large gate
near the surface throws its load well below its centre while a small deep
gate loads almost through its centre. The machinery is the flat-plate
machinery unchanged; only the two shape properties, the area and the second
moment, take the circle's values.

\subsection{A triangular gate, apex up against apex down}

The distribution of a gate's area with depth, not only its centroid, sets
how it loads. Compare two triangular gates of equal area and equal top
submergence, one with its base up and apex down, the other with its base
down and apex up, the pair of Figure~\ref{fig:vol03ch10:triangular-gate}.
Take each as a vertical triangle of base $b = 2\,\si{\meter}$ and height
$H = 3\,\si{\meter}$, with its topmost point at the free surface.

For the \emph{apex-up} gate (base at the top, at the surface; apex $3\,\si{
\meter}$ down), the centroid lies one-third of the height below the base,
\[
h_{C,\text{up}} = \frac{H}{3} = 1.0\,\si{\meter}.
\]
For the \emph{apex-down} gate (apex at the surface; base $3\,\si{\meter}$
down), the centroid lies one-third of the height above the base, that is
two-thirds down from the apex,
\[
h_{C,\text{dn}} = \frac{2H}{3} = 2.0\,\si{\meter}.
\]
Both gates have area $A = \tfrac12 b H = 3\,\text{m}^2$. The resultant
forces scale with the centroidal depth,
\[
F_{\text{up}} = \rho g h_{C,\text{up}} A
= 1000 \times 9.81 \times 1.0 \times 3 \approx 29.4\,\text{kN},
\]
\[
F_{\text{dn}} = \rho g h_{C,\text{dn}} A
= 1000 \times 9.81 \times 2.0 \times 3 \approx 58.9\,\text{kN},
\]
so the apex-down gate carries exactly twice the load of the apex-up gate,
because its centroid sits twice as deep. The lesson is that shape and
orientation, not area alone, decide the hydrostatic load: putting the wide
part of a gate in the deeper water where the pressure is higher increases
both the resultant and the depth of its line of action. A culvert flap, a
weir crest, and a trash rack are all sized with this in mind, and reversing
a triangular gate top-for-bottom doubles the force here without changing a
square metre of area.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-triangular-gate.tex}

\subsection{A hot-air balloon and the temperature it must reach to fly}

A hot-air balloon floats by heating the air inside a fixed-volume envelope
so that the trapped air thins below the density of the surrounding air, the
buoyancy curve of Figure~\ref{fig:vol03ch10:hot-air-balloon}. Take an
envelope of volume $V = 2200\,\text{m}^3$ in ambient air at $\rho_a =
1.225\,\si{\kilogram\per\meter\cubed}$ and $T_a = 288\,\text{K}$, with a
fixed weight (envelope, basket, burner, crew) of $W_{\text{fixed}} =
3500\,\text{N}$. We want the internal air temperature at which the balloon
just lifts off.

Heating the air at constant pressure (the envelope is open at the mouth, so
the inside stays at ambient pressure) thins it in inverse proportion to
absolute temperature, $\rho_i = \rho_a\,T_a/T_i$. The net lift is the
buoyant force (the weight of displaced outside air) minus the weight of the
hot inside air minus the fixed weight,
\[
L = (\rho_a - \rho_i)\,g V - W_{\text{fixed}}
= \rho_a\left(1 - \frac{T_a}{T_i}\right) g V - W_{\text{fixed}}.
\]
Lift-off is $L = 0$. Solving for $T_i$,
\[
1 - \frac{T_a}{T_i} = \frac{W_{\text{fixed}}}{\rho_a g V}
= \frac{3500}{1.225 \times 9.81 \times 2200}
\approx 0.1324,
\]
\[
T_i = \frac{T_a}{1 - 0.1324}
= \frac{288}{0.8676}
\approx 332\,\text{K} \approx 59\,^{\circ}\text{C}.
\]
The balloon lifts off once the inside air reaches about $59\,^{\circ}
\text{C}$, some $44\,^{\circ}\text{C}$ above ambient, and a working balloon
flies a little hotter to carry a payload and to hold a climb, which is why
envelope fabrics are rated to around $120\,^{\circ}\text{C}$ with a
sizeable margin. The physics is the buoyancy balance of a submerged body,
read with the trapped fluid a gas whose density the pilot controls with the
burner: the same $(\rho_a - \rho_i) g V$ that lifts a helium balloon lifts a
hot-air one, but here the density deficit is bought with heat rather than
with a light gas, and it can be turned off by letting the air cool, which is
how the balloon descends.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-hot-air-balloon.tex}

\subsection{A well-type manometer and its area-ratio correction}

A well-type (cistern) manometer replaces one leg of a U-tube with a
large-area reservoir so that a single scale on the narrow tube reads the
pressure, the instrument of Figure~\ref{fig:vol03ch10:well-manometer}. Take
a reservoir of area $A_r = 20\,\text{cm}^2$ and a measuring tube of area
$A_t = 0.5\,\text{cm}^2$, filled with mercury ($\rho = 13\,534\,\si{
\kilogram\per\meter\cubed}$). Applied gauge pressure drives the tube column
up by a reading $R = 80\,\si{\milli\meter}$. We want the true gauge pressure
and the size of the error from ignoring the reservoir drop.

The mercury is incompressible, so the volume that leaves the reservoir
equals the volume that enters the tube, $A_r\,\delta = A_t\,R$, giving the
reservoir drop
\[
\delta = R\,\frac{A_t}{A_r}
= 80 \times \frac{0.5}{20}
= 2.0\,\si{\milli\meter}.
\]
The true height difference of the mercury is the reading plus the reservoir
drop, so the gauge pressure is
\[
p = \rho g\,(R + \delta)
= \rho g\, R\left(1 + \frac{A_t}{A_r}\right)
= 13\,534 \times 9.81 \times 0.080 \times \left(1 + \frac{0.5}{20}\right).
\]
The bare tube reading gives $\rho g R = 13\,534 \times 9.81 \times 0.080
\approx 10.62\,\text{kPa}$, and the area-ratio factor $1 + 0.025$ raises it
to $\approx 10.89\,\text{kPa}$. The reservoir correction is $2.5\,\%$ here,
which a precision instrument removes by one of two means: machine the area
ratio to a known value and print a compressed scale so the tube reads
pressure directly, or make $A_r/A_t$ so large that the correction falls
below the reading resolution. The single-tube convenience of the well
manometer is bought at the price of this small, computable, and
easily-removed bias, and forgetting it biases every reading the same way,
low by the area ratio.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-well-manometer.tex}

\subsection{Wall stress in a submarine pressure hull}

A submarine's cylindrical pressure hull is loaded from outside by the sea,
the reverse of a boiler loaded from inside, and the hydrostatic pressure at
depth turns directly into a compressive hoop stress in the shell. Take a
hull of inner radius $r = 3.5\,\si{\meter}$ and wall thickness $t = 35\,\si{
\milli\meter}$, of steel, at a depth of $d = 300\,\si{\meter}$ in seawater
($\rho_f = 1025\,\si{\kilogram\per\meter\cubed}$). We want the external
pressure and the hoop stress it drives, to see what the hull must survive.

The external pressure is the hydrostatic head at depth,
\[
p = \rho_f g d
= 1025 \times 9.81 \times 300
\approx 3.02 \times 10^6\,\text{Pa}
\approx 3.02\,\text{MPa} \approx 30\,\text{atm}.
\]
For a thin cylindrical shell the hoop stress from a pressure difference $p$
across the wall is $\sigma_\theta = p r/t$ (the thin-wall result carried
over from the stress chapter, here with the pressure acting inward so the
stress is compressive),
\[
\sigma_\theta = \frac{p r}{t}
= \frac{3.02 \times 10^6 \times 3.5}{0.035}
\approx 3.02 \times 10^8\,\text{Pa}
\approx 302\,\text{MPa}.
\]
The hull carries about $300\,\text{MPa}$ of compression in the shell at this
depth, a large fraction of a submarine-steel yield near $550\,\text{MPa}$,
so the diving depth is set with a safety factor against yield and, more
severely, against buckling. A shell in compression fails by buckling well
below the yield stress if it is thin or long, which is why a pressure hull
is stiffened with closely spaced ring frames: the rings shorten the
effective buckling length and raise the collapse pressure far above what the
bare shell could hold. The static pressure calculation gives the load; the
governing check for an externally pressurised hull is the collapse-by-
buckling limit, not the yield of the material, the same lesson the sliding
check taught for the dam, that the weakest mode governs.

\subsection{A hydraulic accumulator and the energy a static pressure stores}

A hydraulic accumulator stores energy by holding oil under pressure against
a compressed gas, releasing it on demand to drive a burst of high flow that
the pump alone could not supply. Take a gas-charged accumulator of total
volume $V_0 = 50\,\si{\liter}$, precharged with nitrogen to $p_0 = 10\,
\text{MPa}$, that is then topped up with oil until the gas is compressed to
$p_1 = 20\,\text{MPa}$. We want the volume of oil stored and the energy
released as the accumulator discharges back to the precharge pressure,
treating the gas compression as isothermal.

The gas obeys $p_0 V_0 = p_1 V_1$ at constant temperature, so at full charge
the gas occupies
\[
V_1 = V_0\,\frac{p_0}{p_1}
= 50 \times \frac{10}{20}
= 25\,\si{\liter},
\]
and the oil stored is the space the gas gave up, $V_{\text{oil}} = V_0 -
V_1 = 25\,\si{\liter}$. The energy released as the gas expands isothermally
from $p_1$ back to $p_0$, pushing the oil out, is the work of expansion
\[
E = \int_{V_1}^{V_0} p\,dV
= p_1 V_1 \ln\!\frac{V_0}{V_1}
= 20 \times 10^6 \times 0.025 \times \ln 2
\approx 5.0 \times 10^5 \times 0.693
\approx 3.47 \times 10^5\,\text{J}.
\]
The accumulator holds about $0.35\,\text{MJ}$, enough to slam a large press
or hold a hydraulic circuit through a pump stall. Two design facts follow
from the arithmetic. The stored oil volume depends on the pressure ratio,
so a higher precharge relative to the working pressure stores less oil but
delivers it at a steadier pressure, and the choice of precharge is a
trade between capacity and flatness of delivery. And the energy is real and
can be dangerous: a charged accumulator is a spring holding hundreds of
kilojoules, and the maintenance discipline is to discharge it to the gas
side before opening the hydraulic circuit, because a line broken under full
charge releases that energy as a jet of oil. The static pressure is not
inert; it is stored work, and the accumulator is the device that makes the
storage explicit.

\subsection{Cabin differential on a pressurised airliner}

An airliner cruising in thin air keeps its cabin at a comfortable pressure,
and the difference between the cabin and the outside air is a static load
the fuselage carries as a pressure vessel. Take an aircraft cruising at
$z = 11\,\si{\kilo\meter}$, where the standard-atmosphere pressure is about
$p_{\text{out}} = 22.6\,\text{kPa}$, holding a cabin pressure equal to the
standard pressure at a cabin altitude of $2400\,\si{\meter}$, about
$p_{\text{cab}} = 75.3\,\text{kPa}$. We want the differential pressure and
the hoop stress it drives in a fuselage of radius $r = 2.0\,\si{\meter}$ and
skin thickness $t = 2.0\,\si{\milli\meter}$.

The cabin differential is the pressure the structure must hold across the
skin,
\[
\Delta p = p_{\text{cab}} - p_{\text{out}}
= 75.3 - 22.6 \approx 52.7\,\text{kPa} \approx 0.53\,\text{atm}.
\]
The fuselage is a thin cylinder loaded from inside, so the hoop stress is
$\sigma_\theta = \Delta p\, r/t$,
\[
\sigma_\theta = \frac{52.7 \times 10^3 \times 2.0}{0.002}
\approx 5.27 \times 10^7\,\text{Pa}
\approx 53\,\text{MPa}.
\]
The skin carries about $53\,\text{MPa}$ of tension around the barrel at
cruise, a modest stress for aluminium alloy in a single application but a
severe one when applied and released on every flight. Each takeoff
pressurises the hull and each landing relieves it, so the cabin
differential is a fatigue load cycled once per flight, and the governing
design limit is the growth of fatigue cracks over tens of thousands of
pressurisation cycles rather than the static strength of the skin. The
static calculation gives the load in one cycle; the number of cycles turns
it into the fatigue problem that sets the safe life and the inspection
interval of a pressurised airframe, the same weakest-mode discipline that
made buckling, not yield, govern the submarine hull.

\subsection{Reading a low pressure on an inclined draft gauge}

A furnace-draft gauge measures a pressure of a few tens of pascals, far too
small for a vertical water column to resolve, so it lays the measuring leg
down at a shallow angle and reads the long slope travel of the meniscus
instead. Take an inclined manometer filled with a light oil of density
$\rho = 820\,\si{\kilogram\per\meter\cubed}$, its tube set at $\theta =
6^{\circ}$ above horizontal, connected to a flue whose draft (a small
suction below atmospheric) draws the oil back along the slope by a read
length $s = 55\,\si{\milli\meter}$ measured along the tube. We want the
gauge pressure and the resolution advantage over a vertical instrument.

The pressure depends on the vertical rise of the fluid, not the slope
travel, and the two are related by the tube angle. The vertical rise for a
slope travel $s$ is
\[
\Delta h = s\,\sin\theta = 0.055 \times \sin 6^{\circ}
\approx 0.055 \times 0.1045 \approx 5.75 \times 10^{-3}\,\si{\meter},
\]
so the gauge pressure magnitude is
\[
|p| = \rho g\,\Delta h
= 820 \times 9.81 \times 5.75 \times 10^{-3}
\approx 46.2\,\text{Pa},
\]
a suction of about $46\,\text{Pa}$ (roughly $4.7\,\si{\milli\meter}$ of
water). The instrument reads this tiny pressure as a comfortable
$55\,\si{\milli\meter}$ of slope travel, an amplification of $s/\Delta h =
1/\sin\theta \approx 9.6$ over the bare vertical rise. The light oil helps
twice over: its low density stretches the vertical rise for a given
pressure, and choosing oil over water avoids the meniscus-pinning and
evaporation troubles of a long thin water thread. The price of the
amplification is a matching sensitivity to levelling: a tube tilted from
its set angle by even a fraction of a degree shifts the zero by the same
$1/\sin\theta$ factor, so an inclined gauge is levelled with a spirit
bubble and zeroed against still air before every reading.

\subsection{Hoop stress at the base of a water tower standpipe}

A municipal water tower holds its supply high so that gravity delivers
pressure to the mains, and the standpipe that carries the column is a thin
cylinder loaded from inside by the head it holds. Take a welded-steel
standpipe of inner radius $r = 3.0\,\si{\meter}$ and wall thickness $t =
10\,\si{\milli\meter}$, filled with water to a height $H = 28\,\si{\meter}$
above its base. We want the internal pressure at the base and the hoop
stress it drives in the wall there, to see what governs the plate
thickness near the bottom.

The internal gauge pressure at the base is the hydrostatic head,
\[
p = \rho g H = 1000 \times 9.81 \times 28
\approx 2.75 \times 10^5\,\text{Pa}
\approx 275\,\text{kPa}.
\]
For a thin cylinder loaded from inside, the hoop stress is $\sigma_\theta =
p r/t$, carried over from the stress chapter,
\[
\sigma_\theta = \frac{p r}{t}
= \frac{2.75 \times 10^5 \times 3.0}{0.010}
\approx 8.24 \times 10^7\,\text{Pa}
\approx 82\,\text{MPa}.
\]
The wall near the base carries about $82\,\text{MPa}$ of tension around the
barrel, a modest fraction of a structural-steel yield near
$250\,\text{MPa}$, so the base course of plate is sized against this hoop
tension with a comfortable factor. Higher up the tank the head is smaller
and the hoop stress falls in step, which is why a tall standpipe is built
with thick plate at the bottom and progressively thinner courses toward the
top: the load is $p r/t = \rho g\,(H - z)\,r/t$, linear in the height of
water still standing above the course in question, and the plate schedule
tapers to match. The same $\rho g H$ that pushes water into the mains sets
the tension the shell must hold, and the two are the same calculation read
for two purposes.

\subsection{A Cartesian diver and the instability of a compressible float}

A Cartesian diver is a small inverted vial trapping an air bubble, floating
in a sealed bottle of water; a squeeze on the bottle sinks it and a release
raises it. The device is a demonstration of the compressibility instability
the submerged-body mastery box named, and it yields to a short calculation.
Take a diver of rigid mass and shell displacing a fixed $V_s = 2.0\,\text{
cm}^3$ of solid, trapping an air bubble that is $V_a = 1.5\,\text{cm}^3$ at
the bottle's resting pressure $p_0 = 101\,\text{kPa}$, with a total hard
mass $m = 3.35\,\si{\gram}$. We want to show it floats at rest and to find
the pressure rise that sinks it.

At rest the diver displaces solid plus bubble, $V_0 = V_s + V_a = 3.5\,
\text{cm}^3$, and the buoyant force is $\rho g V_0 = 1000 \times 9.81
\times 3.5 \times 10^{-6} \approx 3.43 \times 10^{-2}\,\text{N}$, against a
weight of $m g = 3.35 \times 10^{-3} \times 9.81 \approx 3.29 \times
10^{-2}\,\text{N}$. Buoyancy just exceeds weight, so the diver floats,
barely. Squeeze the bottle and the added pressure compresses the trapped
air by Boyle's law, $V_a(p) = V_a\,p_0/p$, shrinking the displacement while
the weight holds fixed. The diver sinks when buoyancy falls to the weight,
\[
\rho g\,[\,V_s + V_a p_0/p\,] = m g,
\]
which solving for the bubble volume needed gives $V_a p_0/p = m/\rho - V_s
= 3.35 \times 10^{-6} - 2.0 \times 10^{-6} = 1.35\,\text{cm}^3$, so the
sinking pressure is
\[
p = p_0\,\frac{V_a}{1.35\,\text{cm}^3}
= 101 \times \frac{1.5}{1.35}
\approx 112\,\text{kPa},
\]
a rise of only about $11\,\text{kPa}$, a tenth of an atmosphere, from a
gentle squeeze. The instability the mastery box described is on display: once
the diver starts down, the growing water pressure compresses the bubble
further, buoyancy falls further, and the descent accelerates, so the diver
does not settle at an intermediate depth but runs to the bottom, and only
releasing the squeeze reverses it. A rigid float in uniform water has no
stable equilibrium depth; the Cartesian diver is that fact made into a toy,
and the same runaway is what a submarine's trim system exists to arrest.

\subsection{Net thrust across a coffer with tailwater on the far side}

The dam and lock examples held water on one side against a dry face. A more
general retaining wall, a cofferdam driven across a river with water partly
backed up on the downstream side, carries the difference of two triangular
prisms, and the net thrust and its line of action come from subtracting one
from the other. Take a sheet-pile coffer wall with fresh water to depth
$H_1 = 5\,\si{\meter}$ on the upstream face and tailwater to $H_2 = 2\,\si{
\meter}$ on the downstream face, per metre of wall. We want the net
horizontal thrust and the height of its line of action above the base.

Each side loads the wall with a triangular pressure prism, resultant
$\tfrac12\rho g H^2$ per metre acting at $H/3$ above the base. Upstream,
\[
P_1 = \tfrac12 \rho g H_1^2
= \tfrac12 \times 1000 \times 9.81 \times 25
\approx 1.23 \times 10^5\,\text{N/m}
\approx 123\,\text{kN/m},
\]
at $H_1/3 \approx 1.67\,\si{\meter}$ above the base; downstream,
\[
P_2 = \tfrac12 \rho g H_2^2
= \tfrac12 \times 1000 \times 9.81 \times 4
\approx 1.96 \times 10^4\,\text{N/m}
\approx 19.6\,\text{kN/m},
\]
at $H_2/3 \approx 0.67\,\si{\meter}$ above the base, pushing the other way.
The net thrust is the difference,
\[
P_{\text{net}} = P_1 - P_2 \approx 123 - 19.6 \approx 103\,\text{kN/m},
\]
directed downstream. Its line of action is found by moment balance about
the base: the net moment is $P_1 (H_1/3) - P_2 (H_2/3) = 123 \times 1.67 -
19.6 \times 0.67 \approx 205 - 13 \approx 192\,\text{kN}\cdot\text{m/m}$,
so the resultant acts at
\[
\bar{y} = \frac{192}{103} \approx 1.87\,\si{\meter}
\]
above the base, higher than the $H_1/3$ point of the upstream prism alone,
because subtracting the low-acting downstream prism lifts the centroid of
what remains. The lesson is that tailwater relieves a retaining wall by the
square of its depth, so even a modest downstream pond cuts the net load
substantially, which is why a lock is analysed with the chamber drained
(the worst case, no relieving tailwater) and why dewatering the downstream
side of a coffer is the event that most raises the load on the sheet piles.

\subsection{Cable tension on a tethered subsurface buoy}

A subsurface buoy holds an instrument package at a fixed depth by floating
against a taut tether anchored to the bottom, the mirror of the crane lift
worked earlier: there the cable held a heavy body up, here it holds a light
body down. Take a spherical buoy of diameter $D = 0.6\,\si{\meter}$ and mass
$m = 45\,\si{\kilogram}$ tethered in seawater ($\rho_f = 1025\,\si{
\kilogram\per\meter\cubed}$), fully submerged. We want the tension the
tether carries.

The buoy's volume is $V = \tfrac{4}{3}\pi (0.3)^3 \approx 0.1131\,\text{m}
^3$, so its buoyant force is
\[
F_B = \rho_f g V
= 1025 \times 9.81 \times 0.1131
\approx 1.14 \times 10^3\,\text{N}
\approx 1.14\,\text{kN},
\]
against a weight of $m g = 45 \times 9.81 \approx 0.44\,\text{kN}$. The
buoyancy exceeds the weight, so the buoy pulls upward and the tether must
carry the difference to hold it down,
\[
T = F_B - m g \approx 1.14 - 0.44 \approx 0.70\,\text{kN}.
\]
The tether runs in tension at about $700\,\text{N}$, and the anchor at the
bottom must weigh (net of its own buoyancy) enough that this uplift cannot
drag it. The tension is independent of the buoy's depth, because a fully
submerged rigid body of fixed volume displaces the same volume at every
depth and its buoyancy does not change with how deep it rides. That
depth-independence is the design convenience of a subsurface mooring: the
tension is set once by the buoy's net buoyancy and does not grow as the
mooring is placed deeper, unlike a surface float whose motion in waves
snaps its line. The one caveat is compressibility, the theme of this run of
examples: a buoy with any trapped gas or a compressible flotation foam
loses a little volume with depth, so its net buoyancy and the tether
tension both sag slightly at greater depth, and a deep mooring is trimmed
with that small loss counted.

\subsection{Pressure across a spinning centrifuge tube}

The rotating-frame pressure field does more than shape a free surface; it
drives the separation inside a filled, capped centrifuge tube, where there
is no free surface at all. Take a sealed tube of liquid in a laboratory
centrifuge, its inner end at radius $r_1 = 3\,\si{\centi\meter}$ and its
outer end at radius $r_2 = 9\,\si{\centi\meter}$ from the spin axis, filled
with a suspension of density $\rho = 1050\,\si{\kilogram\per\meter\cubed}$,
spinning at $\omega = 1000\,\text{rad/s}$. We want the pressure difference
between the outer and inner ends, and the effective gravity that drives the
sedimentation.

In the rotating frame the radial pressure gradient balances the centrifugal
body force, $dp/dr = \rho\,\omega^2 r$, which integrates across the tube to
\[
\Delta p = p(r_2) - p(r_1)
= \tfrac12 \rho\,\omega^2\,(r_2^2 - r_1^2)
= \tfrac12 \times 1050 \times 1000^2 \times (0.09^2 - 0.03^2).
\]
With $r_2^2 - r_1^2 = 0.0081 - 0.0009 = 0.0072\,\text{m}^2$,
\[
\Delta p \approx \tfrac12 \times 1050 \times 10^6 \times 0.0072
\approx 3.78 \times 10^6\,\text{Pa}
\approx 3.8\,\text{MPa},
\]
a pressure difference of nearly four megapascals built across six
centimetres of tube by the spin alone, far larger than the tube's own
gravitational head. The effective gravity at the outer end is $\omega^2 r_2
= 10^6 \times 0.09 = 9.0 \times 10^4\,\text{m/s}^2$, about $9000\,g$, which
is what drives dense particles to the outer end and clarifies the
suspension. The pressure result sizes the tube and the rotor: the closure
and the tube wall must hold several megapascals, and the rotor must contain
the enormous stored rotational energy, which is why a centrifuge rotor is a
guarded, life-limited component whose overspeed failure is a recognised
laboratory hazard. The free-surface paraboloid and this sealed-tube
pressure jump are the same rotating-frame law, $p = p_0 - \rho g z +
\tfrac12\rho\omega^2 r^2$, read once for the shape of an open surface and
once for the load inside a closed tube.

\subsection{When the gas column in a manometer stops being negligible}

Every manometer reduction so far has dropped the light process-gas column
against the dense manometer fluid. The approximation is excellent for a
short gas leg, but a tall one on a high-accuracy reading earns a check.
Take a mercury manometer ($\rho_{\text{Hg}} = 13\,534\,\si{\kilogram\per
\meter\cubed}$) reading a gas pressure through a vertical gas-filled
connecting leg of height $H_g = 3.0\,\si{\meter}$, with the gas at a density
of $\rho_g = 1.6\,\si{\kilogram\per\meter\cubed}$ (a dense process gas near
atmospheric). The mercury height difference is $\Delta h = 200\,\si{
\milli\meter}$. We want the size of the correction the gas column
contributes.

The full path trace from the gas source down the gas leg and across the
mercury gives
\[
p_{\text{gauge}} = \rho_{\text{Hg}} g\,\Delta h - \rho_g g\,H_g.
\]
The mercury term is
\[
\rho_{\text{Hg}} g\,\Delta h
= 13\,534 \times 9.81 \times 0.200
\approx 2.655 \times 10^4\,\text{Pa}
\approx 26.55\,\text{kPa},
\]
and the gas-column correction is
\[
\rho_g g\,H_g
= 1.6 \times 9.81 \times 3.0
\approx 47\,\text{Pa}
\approx 0.047\,\text{kPa}.
\]
The correction is about $0.18\,\%$ of the reading, small but no longer at
the $0.01\,\%$ level of the short-leg gas-line case worked earlier. Whether
it matters is a question the instrument's own accuracy answers: on a
Bourdon gauge good to $\pm 1\,\%$ the gas column is beneath the noise, but
on a precision transducer or a calibration-grade manometer good to
$\pm 0.05\,\%$ the $0.18\,\%$ bias is several times the instrument
uncertainty and must be carried. The rule the example installs is not
``always neglect the gas column'' but ``neglect it only when it falls below
the reading uncertainty,'' and the way to know is to compute both terms,
which the path trace does for free. A dense gas in a tall leg is exactly the
case the habitual approximation gets wrong.

\subsection{The hydrostatic paradox in numbers}

The hydrostatic-paradox mastery box argued that the base force depends on
depth alone, not on the volume of liquid stored. A short calculation makes
the surprise concrete. Take three open vessels, each with the same circular
base of area $A = 0.10\,\text{m}^2$ and each filled with water to the same
height $H = 0.80\,\si{\meter}$: a straight cylinder, a vessel flaring to
twice the base area at the top, and a vessel narrowing to half the base area
at the top. We want the downward force on each base and the weight of water
each holds.

The base force is the base pressure times the base area, and the base
pressure is $\rho g H$ in every case,
\[
F_{\text{base}} = \rho g H\, A
= 1000 \times 9.81 \times 0.80 \times 0.10
\approx 785\,\text{N},
\]
the same $785\,\text{N}$ for all three vessels. The stored weight, however,
differs. The straight cylinder holds $\rho g\,(A H) = 785\,\text{N}$ of
water, matching its base force. The flared vessel holds more, roughly the
average of the base and top areas times the height, about $\rho g\,
[\tfrac12(A + 2A)H] = 1.5 \times 785 \approx 1178\,\text{N}$, half again the
base force; the narrowing vessel holds less, about $\rho g\,[\tfrac12(A +
\tfrac12 A)H] = 0.75 \times 785 \approx 589\,\text{N}$, three-quarters of
it. The base of the flared vessel feels less than the weight it holds
because its sloping walls carry a share downward; the base of the
narrowing vessel feels more than the weight it holds because its inward-
leaning walls pull up on the liquid, and the base makes up the difference.
The number to design to is the base force $\rho g H A$, set by the head, and
reading the base load off the tonnage stored, whether by over- or
under-estimating, is the error the paradox exists to prevent: a flared
settling tank is not loaded by its full contents, and a narrowing standpipe
is loaded by more than it holds.

\subsection{Vignette: face-pressure balance in a slurry tunnel-boring machine}

A modern tunnel-boring machine driving through soft, water-bearing ground
does not simply grind forward and let the face collapse behind the cutter;
it holds the excavated face up with a pressurised fluid, and the pressure
it must hold is a hydrostatic calculation. In a slurry-shield machine the
cutting face is sealed behind a bulkhead, and the chamber between the
cutterhead and the bulkhead is filled with a dense bentonite slurry held at
a controlled pressure. The slurry pushes back on the exposed soil and
groundwater with a pressure that must sit inside a window: high enough to
balance the water pressure and the earth pressure that would otherwise
squeeze the face inward and collapse it, and low enough not to fracture the
ground and blow the slurry out to the surface, a ``blow-out.''

Take a machine boring at a depth where the tunnel axis is $H = 25\,\si{
\meter}$ below the water table, in ground whose pore water is fresh. The
groundwater pressure at the tunnel axis is the hydrostatic head,
\[
p_w = \rho_w g H = 1000 \times 9.81 \times 25 \approx 245\,\text{kPa}.
\]
The slurry must at least balance this, and it must also carry a share of
the earth pressure and a margin for stability, so the target chamber
pressure is set a little above the water pressure, typically the water
pressure plus a support pressure of a few tens of kilopascals. The slurry
itself, being denser than water ($\rho_s \approx 1100$ to $1300\,\si{
\kilogram\per\meter\cubed}$), contributes its own hydrostatic gradient
across the height of the face, so the pressure the machine must hold is
higher at the invert (the bottom of the bore) than at the crown (the top)
by $\rho_s g D$ across a bore of diameter $D$. For a $10\,\si{\meter}$ bore
in slurry at $1200\,\si{\kilogram\per\meter\cubed}$ that gradient is
$1200 \times 9.81 \times 10 \approx 118\,\text{kPa}$ from crown to invert,
a substantial fraction of the mean face pressure, so the machine's
pressure sensors and its control system work to a target that varies over
the height of the face rather than a single number.

The window between collapse and blow-out is what the operator flies the
machine inside, and it narrows as the ground gets weaker and the cover
above the tunnel gets thinner. Too little pressure and the face runs in,
opening a void that can migrate to the surface as a sinkhole; too much and
the slurry fractures the ground and vents, losing face support and
flooding the works. The static-pressure calculation sets the centre of the
window, the water pressure at the axis; the width of the window is a
geotechnical judgment about the ground's strength. The lesson the vignette
carries is that a hydrostatic head, the same $\rho g H$ that loads a dam,
becomes the target a boring machine must actively match, and that matching
a pressure to a computed head is not a static curiosity but a live control
problem on which the safety of the tunnel and the ground above it depends.

\subsection{Vignette: the pressure-balanced float valve and the ballast tank}

A float valve is the quiet piece of fluid statics inside a toilet cistern,
a livestock trough, and a header tank, and the same buoyancy-and-lever
balance scaled up is what trims a submarine. The device is a hollow float
riding on the water surface, linked by a lever arm to a valve that admits
water. As the water level rises, the float rises with it, and the lever
closes the valve; as water is drawn off and the level drops, the float
falls and the valve reopens. The setting is a static balance: at the target
level the buoyant force on the float, acting up through its centre of
buoyancy at the end of the lever, produces a moment about the lever pivot
that just balances the moment the supply pressure exerts on the valve
seat.

Take a spherical float of radius $r = 5\,\si{\centi\meter}$ on a lever arm
of length $L_f = 25\,\si{\centi\meter}$ from the pivot, closing a valve
whose seat area is $A_v = 0.8\,\text{cm}^2$ at a supply pressure of $p_s =
300\,\text{kPa}$, with the valve stem $L_v = 3\,\si{\centi\meter}$ from the
pivot. The force the supply pressure exerts on the closed valve is
\[
F_v = p_s A_v = 300 \times 10^3 \times 0.8 \times 10^{-4}
\approx 24\,\text{N},
\]
and to hold the valve shut the float must supply a moment $F_v L_v = 24
\times 0.03 \approx 0.72\,\text{N}\cdot\text{m}$ about the pivot. The float
supplies it by buoyancy, and the buoyant force needed at the float,
$F_b = F_v L_v/L_f = 0.72/0.25 \approx 2.9\,\text{N}$, is the weight of
water the float must displace, about $0.29\,\si{\kilogram}$ or roughly a
third of the float's $524\,\text{cm}^3$ volume submerged. The float rides
partly submerged at the closing point, and the depth of submersion adjusts
the closing moment, so the level at which the valve shuts is set by the
float geometry and the lever ratio, the numbers a plumber trims by bending
the arm.

The same balance, scaled and inverted, trims a submarine. A submarine
adjusts its buoyancy by admitting water into or blowing air out of ballast
tanks: flooding a tank raises the vessel's weight until it exceeds
buoyancy and the boat sinks, blowing the tank with compressed air restores
buoyancy and it rises, and neutral trim is the static balance where weight
equals the buoyant force of the fully submerged hull. The trim is a
knife-edge because a fully submerged hull, unlike a floating one, has no
waterplane to supply a restoring force against a change of depth: displace
it downward and its displacement does not grow, so a submerged submarine at
neutral buoyancy is in neutral, not stable, vertical equilibrium and must
trim continuously against changes in water density and hull compression.
The float valve and the ballast tank are the same physics at opposite
scales, buoyancy set against a load through a linkage, and both are read
off the static balance of displaced weight against the force to be held.

\subsection{Vignette: the hydraulic intensifier and the enclosed press}

A hydraulic press exchanges force for distance through the area ratio of two
pistons, the balance drawn in Figure~\ref{fig:vol03ch10:hydraulic-press},
and the same trick chained twice builds a machine that raises a pump's
pressure to a level the pump could not reach alone. A workshop press might
have a pump that comfortably delivers oil at $20\,\text{MPa}$, while a
cold-forming or die-testing operation needs $200\,\text{MPa}$ at the tool.
The intensifier bridges the gap. It is a single sliding assembly with a
large piston on the low-pressure side and a small piston on the
high-pressure side, rigidly joined, so the pump pressure acting on the large
face is balanced by a much higher pressure on the small face. Force balance
on the moving assembly reads $p_{\text{low}} A_{\text{large}} =
p_{\text{high}} A_{\text{small}}$, so the pressure gain is the piston-area
ratio,
\[
\frac{p_{\text{high}}}{p_{\text{low}}}
= \frac{A_{\text{large}}}{A_{\text{small}}}.
\]
An intensifier with a large-to-small area ratio of ten turns a
$20\,\text{MPa}$ pump into a $200\,\text{MPa}$ supply at the tool, at the
cost of delivering only a tenth of the pump's volume flow, the same
force-for-distance bargain the press pays, now spent on pressure rather than
force. The static reasoning is Pascal's principle applied to a closed
assembly: pressure transmits undiminished through each fluid body, and the
step up happens at the mechanical joint where two areas share one force, not
in the fluid itself.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-hydraulic-press.tex}

The design cautions are those of any high-pressure static system carrying
stored energy. The high-pressure chamber holds oil at a pressure where the
liquid's own compressibility, a few tenths of a per cent per ten megapascals,
becomes a real volume that must be pumped before any tool motion begins, so
a stiff high-pressure circuit with the air bled out is not a nicety but a
requirement for a press that responds to its control. And the stored elastic
energy in the compressed oil and the strained steel is released violently if
a high-pressure line lets go, so the discipline that governs a hydraulic
accumulator governs an intensifier: relieve the pressure to the low side
before opening the circuit, and rate every fitting on the small-piston side
for the intensified pressure, not the pump pressure, because the whole point
of the machine is that the two are not the same number.

\subsection{Vignette: the standpipe test and why plumbers pressurise with water}

Before a building's drain and vent piping is closed into the walls, it is
proved tight by filling the stack with water and watching for a drop, a test
whose severity is pure hydrostatics. A plumber caps the base of a vertical
stack and fills it from the top through a temporary standpipe, building a
column of water whose head loads every joint below. A ten-storey stack
filled to the roof puts perhaps $30\,\si{\meter}$ of water head on the
lowest joints, so the pressure there is
\[
p = \rho g H = 1000 \times 9.81 \times 30
\approx 2.94 \times 10^5\,\text{Pa}
\approx 294\,\text{kPa},
\]
about three atmospheres, a load the finished system will never see in normal
drainage service, where the pipe runs mostly empty and at atmospheric
pressure. The test is deliberately harsher than service, and that is its
value: a joint that holds three atmospheres of static head will not weep
under the trickle of ordinary use, and a joint that fails the test is found
now, open to view, rather than later behind a finished wall.

Two features of the test are hydrostatic lessons in plain clothes. The
pressure depends only on the height of the column, not on the pipe diameter
or the volume of water, so a narrow stack and a wide one filled to the same
height test their bottom joints to the same pressure, the depth-alone rule
of the hydrostatic paradox seen from the plumber's end. And the test is run
in sections on a tall building precisely because the head at the base of a
full-height fill would overload the lowest fittings, so the stack is proved
a few storeys at a time, each section seeing only its own column. The
temporary standpipe that gives the test its name is a manometer read for its
absolute height rather than a difference: the plumber sets the head by the
height of water above the joint under test, and reads a failure as a falling
level over time. The same $\rho g H$ that sizes a dam and lifts a water
tower proves a drain line tight, one number serving structure, supply, and
inspection alike.

\begin{mastery}
\textbf{Pressure in a static fluid of depth-varying density: ocean and
atmosphere by the same integral.} The incompressible law $p=\rho gh$
assumes a constant density. Take instead a fluid whose density varies with
depth, $\rho=\rho(z)$, still in hydrostatic balance $dp/dz=-\rho(z)g$ with
$z$ measured upward. The pressure is the running integral of the local
specific weight,
\[
p(z) = p_0 + g\int_z^{z_0}\rho(z')\,dz',
\]
which is the layered-tank sum of an earlier worked example, taken to the
continuous limit: every thin slab $dz'$ adds its own $\rho g\,dz'$, and the
total is the area under the density profile. Two cases close the chapter's
arc. In the deep ocean the density rises with depth because seawater
compresses under load; writing $\rho=\rho_0(1+p/K)$ with $K$ the bulk
modulus turns the balance into $dp/dz=-\rho_0 g(1+p/K)$, whose solution
grows slightly faster than linear, the few-per-cent correction the Mariana
Trench estimate carried. In the atmosphere the density falls with height
because the gas expands as pressure drops; with $\rho=pM/(RT)$ at constant
temperature the same integral gives the exponential barometric law
$p=p_0\exp(-z/H)$, and with a linear temperature profile it gives the
power-law standard-atmosphere segments. One integral, $p=p_0+g\int\rho\,
dz$, covers the incompressible liquid (constant $\rho$, linear $p$), the
slightly compressible ocean (rising $\rho$, faintly super-linear $p$), and
the compressible atmosphere (falling $\rho$, exponential $p$), the three
shapes plotted together in Figure~\ref{fig:vol03ch10:density-laws}. The
discipline the chapter has built, identify the element, write the balance,
integrate, is the same in all three; only the density law inside the
integral changes.
\end{mastery}

\input{volumes/03-force/ch10-fluid-statics/figures/fig-density-laws.tex}

\begin{mastery}
\textbf{Why a submerged body has no vertical stability, and where a
neutrally buoyant one settles.} A floating body draws its vertical
stability from its waterplane: push it down and the extra displaced volume
of the freshly immersed sidewall adds buoyancy that pushes it back up, the
stiffness the metacentric analysis measures. A fully submerged body has no
waterplane, so this restoring stiffness is gone, and its vertical
equilibrium is governed instead by how its own density compares with the
surrounding fluid's at each depth. Take a rigid body of fixed volume and
weight in a fluid whose density rises with depth, the stable stratification
of the ocean below the mixed layer. At the depth where the fluid density
equals the body's mean density the net vertical force is zero. Displace the
body downward a little: the fluid around it is now denser than the body, so
buoyancy exceeds weight and pushes it back up. Displace it upward: the
fluid is lighter, weight wins, and it sinks back. The density gradient
supplies the restoring force that the missing waterplane cannot, and the
equilibrium is stable, a depth the body seeks and holds. This is the
principle of the isopycnal float and of the neutrally buoyant
oceanographic drifter, which is trimmed to settle on a chosen density
surface and ride it for months. The catch is that a compressible body,
one that shrinks under pressure faster than the water does, has the sign
reversed: pushed down it loses volume, loses buoyancy, and runs away
downward, which is the instability a submarine fights, because its hull
compresses slightly with depth and its trim is therefore a knife-edge that
must be tended continuously against both the rising water density and its
own shrinking displacement. Vertical stability underwater is a contest
between the fluid's density gradient and the body's own compressibility,
and which one wins decides whether a depth is sought or fled.
\end{mastery}

\begin{mastery}
\textbf{The angle of loll: stable equilibrium at a heel.} A vessel whose
upright metacentric height is slightly negative does not necessarily
capsize. As it heels, the waterplane shape changes, and for a hull with
flared sides or a rising bilge the second moment $I_{\text{wp}}$ of the
heeled waterplane grows with angle. The metacentric radius
$\overline{BM} = I_{\text{wp}}/V_{\text{disp}}$ rises with it, so the
effective metacentric height, negative at upright, can climb back through
zero at some finite heel. At that angle the righting arm vanishes with a
positive slope: the vessel has found a stable equilibrium tilted to one
side, the \engterm{angle of loll}. A lolling vessel sits stubbornly heeled,
flopping from one side to the other as it rolls through upright, and the
condition is dangerous precisely because it looks like a simple list while
the reserve of stability is small and the deck edge is closer to the water.
The cure is to raise the true metacentric height, by lowering weight or by
pressing up slack tanks to kill free surface, not by shifting weight to the
high side, which only swaps the loll to the other beam. The angle of loll
is the floating-body counterpart of a buckled column that finds a stable
bent shape: the symmetric state has gone unstable, and the system settles
into a tilted one that is locally stable but globally fragile.
\end{mastery}

\section{End-to-end case study: stability of a concrete gravity dam}\pathtag{standard}

The chapter's separate results, hydrostatic thrust, centre of pressure,
self-weight, and the moment balance, meet in a single calculation when an
engineer asks the one question a dam must answer: will it stand? We carry
a concrete gravity dam through the four checks that decide the answer,
overturning, sliding, base pressure, and the resulting factors of safety,
keeping numbers throughout. The geometry is the free body of
Figure~\ref{fig:vol03ch10:dam-fbd}.

\subsection{The structure and the loads}

Take a non-overflow concrete gravity dam of height $H = 40\,\si{\meter}$,
with a vertical upstream face and a battered downstream face, giving a
trapezoidal cross-section: crest width $a = 6\,\si{\meter}$, base width
$B = 30\,\si{\meter}$. The reservoir is full to the crest. We work per
metre of dam length ($w = 1\,\si{\meter}$ into the page), the standard
unit for a gravity-dam section. Concrete unit weight is $\gamma_c =
\rho_c g = 2400 \times 9.81 \approx 23.5\,\text{kN/m}^3$; water unit
weight is $\gamma_w = 998 \times 9.81 \approx 9.79\,\text{kN/m}^3$.

\emph{Horizontal hydrostatic thrust.} The upstream face is vertical, so
the thrust is the triangular pressure prism,
\[
P_h = \tfrac12 \gamma_w H^2
= \tfrac12 \times 9.79 \times 40^2
\approx 7.83 \times 10^3\,\text{kN}
= 7830\,\text{kN per metre},
\]
acting horizontally at $H/3 = 13.33\,\si{\meter}$ above the base.

\emph{Self-weight.} Split the trapezoid into a rectangle ($a \times H$)
and a triangle (base $B - a = 24\,\si{\meter}$, height $H$). The
rectangle area is $6 \times 40 = 240\,\text{m}^2$, weight $W_1 =
\gamma_c \times 240 \approx 5640\,\text{kN}$, centroid at $a/2 = 3\,
\si{\meter}$ from the upstream (vertical) face. The triangle area is
$\tfrac12 \times 24 \times 40 = 480\,\text{m}^2$, weight $W_2 =
\gamma_c \times 480 \approx 11\,280\,\text{kN}$, centroid at $a + (B -
a)/3 = 6 + 8 = 14\,\si{\meter}$ from the upstream face. Total weight
$W = W_1 + W_2 \approx 16\,920\,\text{kN}$ per metre.

\emph{Uplift.} Pore water under the base pushes up. With a drained
foundation we take a linear uplift from full reservoir head at the heel
to zero at the toe (a conservative-enough textbook assumption; real
designs apply a drainage-gallery reduction). The uplift pressure at the
heel is $\gamma_w H = 9.79 \times 40 \approx 392\,\text{kPa}$, falling
linearly to zero at the toe, so the uplift resultant is the triangular
prism
\[
U = \tfrac12 (\gamma_w H)(B)
= \tfrac12 \times 392 \times 30
\approx 5870\,\text{kN per metre},
\]
acting upward at $B/3 = 10\,\si{\meter}$ from the heel, that is $20\,
\si{\meter}$ from the toe.

\subsection{Overturning check}

Take moments about the downstream toe $O$. The overturning moment is the
horizontal thrust times its lever arm above the base plus the uplift
times its lever arm to the toe; the resisting moment is the self-weight
times its lever arm to the toe.
\[
M_{\text{over}} = P_h \times \frac{H}{3} + U \times \frac{B}{3}
= 7830 \times 13.33 + 5870 \times 10
\approx 104\,400 + 58\,700
\approx 163\,100\,\text{kN}\cdot\text{m}.
\]
The resisting moment uses each weight's distance from the toe (the toe is
at $B = 30\,\si{\meter}$ from the upstream face): $W_1$ at $30 - 3 = 27\,
\si{\meter}$, $W_2$ at $30 - 14 = 16\,\si{\meter}$:
\[
M_{\text{resist}} = W_1 \times 27 + W_2 \times 16
= 5640 \times 27 + 11\,280 \times 16
\approx 152\,300 + 180\,500
\approx 332\,800\,\text{kN}\cdot\text{m}.
\]
The factor of safety against overturning is
\[
\mathrm{FS}_{\text{over}} = \frac{M_{\text{resist}}}{M_{\text{over}}}
= \frac{332\,800}{163\,100}
\approx 2.04.
\]
A gravity dam is normally required to carry $\mathrm{FS}_{\text{over}}
\ge 1.5$ for the usual load combination, so the section passes
overturning with margin to spare.

\subsection{Sliding check}

Sliding resistance is the friction the base can mobilise against the
horizontal thrust. The normal force on the base is the self-weight
reduced by the uplift,
\[
N = W - U = 16\,920 - 5870 \approx 11\,050\,\text{kN per metre}.
\]
With a concrete-on-rock friction coefficient $\mu = 0.70$ (a typical
design value; a real design also credits foundation cohesion), the
available friction is $\mu N = 0.70 \times 11\,050 \approx 7740\,\text{
kN}$, and the sliding factor of safety is
\[
\mathrm{FS}_{\text{slide}} = \frac{\mu N}{P_h}
= \frac{7740}{7830}
\approx 0.99.
\]
The section fails sliding: the available friction barely equals the
driving thrust, well below the $\mathrm{FS}_{\text{slide}} \ge 1.5$ a
design would demand. Uplift is the culprit; it removes a third of the
normal force and so a third of the friction. The remedy a designer
reaches for is a drainage gallery and a grout curtain near the heel,
which cut the uplift roughly in half. Recomputing with $U' \approx
2940\,\text{kN}$ gives $N' = 16\,920 - 2940 \approx 13\,980\,\text{kN}$,
$\mu N' \approx 9790\,\text{kN}$, and
\[
\mathrm{FS}_{\text{slide}}' = \frac{9790}{7830} \approx 1.25,
\]
still short of $1.5$. The section is too light; either the base must be
widened or a foundation key (a shear key cut into the rock) must be
added. The sliding check, not overturning, governs this geometry, which
is the usual finding for a gravity dam on a competent foundation: a dam
that is safe against tipping can still be on the edge of sliding once
uplift is counted.

\subsection{Base-pressure (no-tension) check}

The base pressure must stay compressive across the whole base, because
the concrete-rock joint cannot carry tension. The check is whether the
resultant of all vertical forces falls within the middle third of the
base (the kern). Locate the resultant by taking moments of the vertical
forces about the toe and dividing by the net vertical force. The net
moment about the toe of the vertical forces alone, plus the horizontal
thrust's moment, fixes the position $\bar{x}$ of the resultant from the
toe:
\[
\bar{x} = \frac{M_{\text{resist}} - M_{\text{over}}}{N}
= \frac{332\,800 - 163\,100}{11\,050}
\approx 15.4\,\si{\meter} \text{ from the toe.}
\]
The eccentricity from the base centre ($B/2 = 15\,\si{\meter}$ from the
toe) is $e = 15.4 - 15.0 = 0.4\,\si{\meter}$, which is inside the kern
limit $B/6 = 5\,\si{\meter}$. The base pressures at the toe and heel are
\[
p_{\text{toe, heel}} = \frac{N}{B}\left(1 \pm \frac{6e}{B}\right)
= \frac{11\,050}{30}\left(1 \pm \frac{6 \times 0.4}{30}\right)
= 368 \times (1 \pm 0.08),
\]
so $p_{\text{toe}} \approx 398\,\text{kPa}$ and $p_{\text{heel}} \approx
339\,\text{kPa}$, both compressive. The base-pressure check passes, and
the small eccentricity confirms the section is far from the tension limit;
the distribution is the gentle trapezoid of
Figure~\ref{fig:vol03ch10:dam-basepressure}, not the disallowed
heel-tension case.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-dam-fbd.tex}

\input{volumes/03-force/ch10-fluid-statics/figures/fig-dam-basepressure.tex}

\subsection{Verdict}

The section is comfortable against overturning ($\mathrm{FS} \approx
2.0$) and against base tension (eccentricity $0.4\,\si{\meter}$ inside a
$5\,\si{\meter}$ kern), but it does not meet the sliding criterion once
uplift is included, even after foundation drainage. The honest verdict is
that the trial geometry must change: widen the base, batter the upstream
face to add a stabilising wedge of water weight on the dam, or cut a
shear key. The case study makes the discipline visible. A single
favourable check proves nothing; the governing check is the one with the
lowest factor of safety, and for a gravity dam that check is usually
sliding under uplift. The same four checks, with the same hydrostatic
thrust and the same moment balance, size every gravity retaining
structure, from a reservoir dam to a quay wall to a basement retaining
wall holding back saturated soil.

\begin{principle}[The governing check is the weakest one]
A structure holding back fluid passes only if every check passes:
overturning, sliding, and base pressure each carry their own factor of
safety, and the design is governed by whichever is smallest. Uplift on
the base attacks sliding and base pressure together by removing normal
force, so a fluid-retaining structure that looks safe on weight and tip
alone can still be unsafe once pore pressure under the base is counted.
The discipline is to run all the checks and report the minimum, not the
most reassuring.
\end{principle}

\section{End-to-end case study: stability of a loaded pontoon}\pathtag{standard}

The dam case study carries the static-thrust results to a verdict on a
fixed structure. The buoyancy and metacentric results carry to a verdict
on a floating one, and the calculation has a different texture: the
unknown is no longer a factor of safety against a single failure surface
but a metacentric height that must stay positive across the whole working
range of loading. We carry a rectangular work pontoon through draft,
centre of buoyancy, metacentric height, the righting moment it produces,
and the way a partly filled tank on deck erodes the margin. The numbers
run end to end; the section of the pontoon is
Figure~\ref{fig:vol03ch10:pontoon-section}.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-pontoon-section.tex}

\subsection{The vessel and the light condition}

Take a flat-bottomed steel pontoon, length $L = 18\,\si{\meter}$, beam
$B = 8\,\si{\meter}$, depth (keel to deck) $D = 2.0\,\si{\meter}$, floating
in fresh water ($\rho_f = 1000\,\si{\kilogram\per\meter\cubed}$). The empty
hull (the light ship) has mass $m_0 = 36\,\si{\tonne}$ with its centre of
gravity at $\overline{KG}_0 = 0.9\,\si{\meter}$ above the keel, a value a
flat-bottomed hull earns by carrying most of its steel low in the bottom
and side plating. The pontoon is to carry a deck load of $m_c =
60\,\si{\tonne}$ whose centre sits $1.8\,\si{\meter}$ above the keel.

\emph{Draft in the loaded condition.} Vertical equilibrium sets the
displaced volume equal to the total mass divided by the water density. The
total mass is $m = m_0 + m_c = 96\,\si{\tonne}$, so
\[
V_{\text{disp}} = \frac{m}{\rho_f} = \frac{96{,}000}{1000}
= 96\,\text{m}^3,
\qquad
T = \frac{V_{\text{disp}}}{L B} = \frac{96}{18 \times 8}
\approx 0.667\,\si{\meter}.
\]
The pontoon floats at two-thirds of a metre draft, leaving freeboard
$D - T \approx 1.33\,\si{\meter}$. The draft is shallow because the
waterplane is large: a wide flat hull buys low draft, which is the reason
work barges are built broad and shallow rather than deep and narrow.

\subsection{Centre of buoyancy and metacentric height}

For a wall-sided hull the underwater body is a rectangular block, so the
centre of buoyancy sits at half the draft above the keel,
\[
\overline{KB} = \frac{T}{2} \approx 0.333\,\si{\meter}.
\]
The waterplane is the full deck rectangle, $L \times B$, and its second
moment about the longitudinal roll axis is
\[
I_{\text{wp}} = \frac{L B^3}{12} = \frac{18 \times 8^3}{12}
= 768\,\text{m}^4,
\]
which gives the metacentric radius
\[
\overline{BM} = \frac{I_{\text{wp}}}{V_{\text{disp}}}
= \frac{768}{96} = 8.0\,\si{\meter}.
\]
The combined centre of gravity is the mass-weighted average of the hull
and the cargo heights,
\[
\overline{KG} = \frac{m_0\,\overline{KG}_0 + m_c\,\overline{KG}_c}{m}
= \frac{36 \times 0.9 + 60 \times 1.8}{96}
= \frac{32.4 + 108}{96}
\approx 1.46\,\si{\meter}.
\]
The metacentric height follows from the keel-referenced sum,
\[
\overline{GM} = \overline{KB} + \overline{BM} - \overline{KG}
= 0.333 + 8.0 - 1.46
\approx 6.87\,\si{\meter}.
\]
The value is large and comfortably positive: the broad waterplane gives a
metacentric radius of eight metres, which dwarfs the modest height of the
loaded centre of gravity. The pontoon is stiff in roll, so stiff that a
naval architect would worry about the opposite problem, a short snappy roll
period that is hard on deck crew and cargo lashings, rather than about a
stability deficit.

\subsection{Righting moment and the working heel}

The small-angle righting arm is $\overline{GZ} = \overline{GM}\sin\theta$,
so the righting moment at a heel angle $\theta$ is
\[
M_r(\theta) = m g\,\overline{GM}\sin\theta.
\]
A wind gust or an off-centre lift might apply a heeling moment. Suppose the
crane on deck lifts a $5\,\si{\tonne}$ load at a transverse offset of
$3\,\si{\meter}$ from the centreline. The heeling moment is
\[
M_h = m_{\text{lift}}\, g\, d
= 5000 \times 9.81 \times 3
\approx 1.47 \times 10^5\,\text{N}\cdot\text{m}.
\]
The pontoon heels until the righting moment matches it. Setting
$M_r = M_h$ and solving for the angle, with $m g = 96{,}000 \times 9.81
\approx 9.42 \times 10^5\,\text{N}$,
\[
\sin\theta = \frac{M_h}{m g\,\overline{GM}}
= \frac{1.47 \times 10^5}{9.42 \times 10^5 \times 6.87}
\approx 0.00227,
\qquad
\theta \approx 0.13^{\circ}.
\]
The steady heel from the off-centre lift is barely a tenth of a degree,
the visible sign of the large metacentric height: a stiff hull shrugs off
a heeling moment with a tiny heel.

\subsection{The free-surface effect that erodes the margin}

The comfortable margin is real only while the deck load is solid. Replace
part of the cargo with a partly filled water tank, $4\,\si{\meter}$ long
and $6\,\si{\meter}$ wide, carried on deck. The tank's free surface has a
second moment about its own longitudinal axis of
\[
i = \frac{\ell\, b^3}{12} = \frac{4 \times 6^3}{12} = 72\,\text{m}^4,
\]
and the partly filled liquid behaves, for stability, as though the centre
of gravity rose by a virtual amount
\[
\overline{GG'} = \frac{\rho_l\, i}{\rho_f\, V_{\text{disp}}}
= \frac{1000 \times 72}{1000 \times 96}
= 0.75\,\si{\meter},
\]
taking the tank liquid as water ($\rho_l = \rho_f$). The free-surface rise
does not depend on how much liquid the tank holds, only on the area of its
surface; a tank an inch deep and a tank nearly full give the same loss as
long as a free surface exists. The effective metacentric height drops to
\[
\overline{GM}_{\text{eff}} = \overline{GM} - \overline{GG'}
\approx 6.87 - 0.75 \approx 6.12\,\si{\meter}.
\]
On this stiff pontoon the loss of three-quarters of a metre is a dent, not
a danger. The same loss on a hull whose bare $\overline{GM}$ was already
under a metre, a tall vessel or a heavily laden one with cargo high,
turns a positive metacentric height negative and the vessel rolls over at
the dock. The mechanism is the geometry of
Figure~\ref{fig:vol03ch10:free-surface}: the liquid migrates to the low
side as the hull heels, and the migrating wedge does exactly what a slack
weight on a swinging line would do, supplying no restoring effect and
stealing the support that would otherwise resist the heel.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-free-surface.tex}

\subsection{Verdict}

The pontoon floats at $0.67\,\si{\meter}$ draft, carries its design load
with a metacentric height near seven metres, heels a tenth of a degree
under a realistic off-centre lift, and keeps a metacentric height above
six metres even with a free-surface tank on deck. The verdict is a stable,
indeed a stiff, vessel, and the design lesson is the inverse of the dam
case: there the governing number was a factor of safety that came out too
low; here the governing number is a metacentric height that comes out
generously high, with the only caution that the free-surface effect, not
the solid load, is the term most likely to surprise an operator who adds a
tank of liquid without recomputing the stability. The same five
quantities, draft, centre of buoyancy, metacentric radius, metacentric
height, and free-surface correction, decide the stability of every
floating structure, from this work pontoon to a semi-submersible drilling
rig to a loaded car ferry.

\begin{principle}[Free surface steals stability regardless of quantity]
A partly filled tank reduces a floating body's effective metacentric
height by $\rho_l\, i/(\rho_f V_{\text{disp}})$, with $i$ the second moment
of the free surface, a loss that depends on the surface area of the liquid
and not on how much liquid the tank holds. A nearly empty slack tank is as
damaging as a half-full one. The remedy is to subdivide the tank with
baffles, cutting $i$ as the cube of each cell's width, or to press the tank
full so that no free surface remains. Every stability calculation on a
vessel that carries liquids must count the free-surface correction in the
worst slack condition, not the convenient full one.
\end{principle}

\section{End-to-end case study: a canal lock mitre gate}\pathtag{standard}

The dam case study held back a static head with a fixed mass; the pontoon
case study floated a load and watched a metacentric height. A mitre lock
gate holds back a head too, but it carries the load in a way neither of
those does: a pair of leaves leans into the high water so that the
hydrostatic thrust wedges them shut, and the load travels out of each leaf
as a compression along its own length into the masonry at the far end. We
carry one leaf of a mitre pair through the thrust on its face, the centre
of pressure, the mitre thrust at the meeting point, and the quoin reaction
at the hinge, keeping numbers throughout. The plan geometry is
Figure~\ref{fig:vol03ch10:lock-gate}.

\input{volumes/03-force/ch10-fluid-statics/figures/fig-lock-gate.tex}

\subsection{The gate and the head}

A lock chamber of clear width $W = 8\,\si{\meter}$ is closed by a pair of
mitre gates against an upstream water depth $H = 6\,\si{\meter}$, with the
downstream side drained dry for the calculation (the worst case for the
gates, since any tailwater would relieve part of the load). Each leaf spans
half the chamber and meets its partner at the mitre point on the
centreline. The leaves lean into the high water at a mitre half-angle
$\beta = 20^{\circ}$ measured between a leaf and the line straight across
the chamber, the angle that sets how hard the head wedges the pair shut.
Fresh water has specific weight $\gamma_w = \rho g = 998 \times 9.81
\approx 9.79\,\text{kN/m}^3$.

\emph{Length of a leaf.} Each leaf reaches from its quoin at the wall to
the mitre point on the centreline, a horizontal run of $W/2 = 4\,\si{
\meter}$ across the chamber. With the leaf set back at the mitre angle, its
true length along the face is
\[
\ell = \frac{W/2}{\cos\beta} = \frac{4}{\cos 20^{\circ}}
\approx \frac{4}{0.940} \approx 4.26\,\si{\meter}.
\]
The wetted face of one leaf is therefore a vertical rectangle of width
$\ell \approx 4.26\,\si{\meter}$ and height $H = 6\,\si{\meter}$.

\subsection{Thrust on one leaf and its centre of pressure}

The water stands to the top of the gate, so the pressure on the leaf face
is the familiar triangular prism, zero at the surface and $\rho g H$ at the
sill. The resultant normal force on one leaf is the pressure at the
centroid times the area,
\[
P = \rho g\,(H/2)\,(\ell H)
= 9.79 \times 10^3 \times 3 \times (4.26 \times 6)
\approx 7.51 \times 10^5\,\text{N}
\approx 751\,\text{kN},
\]
acting horizontally, normal to the leaf face. Its line of action sits at
the centroid of the triangular pressure prism, at $H/3 = 2.0\,\si{\meter}$
above the sill. The vertical position is what the gate's horizontal framing
(the strongbacks and the bottom girder) must be sized for; the horizontal
position, along the leaf, sits at the mid-width of the rectangle, $\ell/2
\approx 2.13\,\si{\meter}$ from the quoin, because the pressure is uniform
across the width at any one depth.

\subsection{Mitre thrust and quoin reaction}

A single leaf is a two-force-plus-load member in plan: the distributed
water thrust $P$ on its face, a contact reaction at the mitre point $M$
where it presses against its partner, and the reaction at the quoin post
$Q$ where it bears on the masonry. Both end reactions act along the leaf to
first order, the mitre contact as a compression pushing back along the leaf
and the quoin as the equal and opposite bearing thrust. Resolve the water
load. The thrust $P$ acts normal to the leaf; the leaf makes the angle
$\beta$ with the cross-chamber line, so the component of $P$ along the leaf
direction is what the end reactions carry. Taking moments and force balance
on the leaf in plan, the compressive thrust transmitted through the mitre
contact is
\[
T_m = \frac{P}{2\sin\beta}
= \frac{751}{2 \times \sin 20^{\circ}}
= \frac{751}{2 \times 0.342}
\approx 1.10 \times 10^3\,\text{kN}
= 1.10\,\text{MN}.
\]
The mitre thrust is larger than the water load that produces it, by the
factor $1/(2\sin\beta) \approx 1.46$ at this angle. The quoin reaction has
the same magnitude along the leaf by force balance, $R_q \approx T_m
\approx 1.10\,\text{MN}$, directed from the mitre back into the hinge post,
with a transverse component that the heel of the gate transfers into the
quoin recess.

The thrust multiplication is the design tension at the heart of a mitre
gate. A shallow mitre (small $\beta$) wedges the leaves tightly and lets a
slim gate hold a large head, but it drives the mitre and quoin thrust up as
$1/\sin\beta$, so the masonry bearing and the gate's own compression
capacity climb steeply as the angle is reduced. A square gate ($\beta \to
90^{\circ}$) would carry the smallest end thrust, $T_m = P/2$, but it could
not wedge shut at all and would have to be held by its hinges alone against
the full head. The classical mitre angle of about $18^{\circ}$ to
$22^{\circ}$, the value used since the canals of the seventeenth and
eighteenth centuries, balances a manageable thrust against a gate slim
enough to swing.

\subsection{Verdict}

One leaf of this pair carries a water thrust of about $751\,\text{kN}$,
delivered into the structure as a mitre compression and a quoin bearing
reaction near $1.1\,\text{MN}$ each, acting at $H/3$ above the sill. The
governing checks for the gate are the compression capacity of the leaf as a
horizontal arch or beam-column carrying $T_m$, the bearing stress at the
quoin and at the mitre contact (a stress concentration the mitre seal must
also accommodate), and the watertightness of the bottom sill seal where the
pressure is highest. The same plan-force resolution sizes every mitre or
pointing gate, from a narrow-boat lock to the largest sea lock, and it
shows why the gates point into the high water: the head that loads them is
the same head that holds them shut. A gate hung the wrong way round would
be blown open by the very pressure a correctly pointed pair turns into a
sealing thrust.

\begin{principle}[A pointing gate turns its load into its seal]
A mitre gate set at a half-angle $\beta$ into the high water carries a face
thrust $P=\rho g(H/2)(\ell H)$ as a compression $T_m=P/(2\sin\beta)$ through
the mitre and quoin. The end thrust exceeds the water load by $1/(2\sin
\beta)$, so a shallow mitre buys a slim gate at the cost of high bearing
stress, and the same head that loads the leaves is the head that wedges them
shut. The discipline is to point the gate into the head, to check the end
thrust and bearing rather than the face load alone, and to seal hardest
where the pressure is greatest, at the sill.
\end{principle}

\section{Failure: dam breaches and floating instability}\pathtag{core}

Fluid-statics failures cluster into two families: failures of
structures holding back static fluid (dams, retaining walls,
tank walls, ship hulls) and failures of floating bodies whose
stability has been lost (capsizing, sinking, list-induced
failure of equipment). The chapter's failure section names the
canonical mechanisms.

\subsection{Dam-breach mechanisms}

A dam is a structure designed to hold back a static head of
water against the hydrostatic pressure on its upstream face. The
hydrostatic load on a dam of height $H$ is
\[
F = \tfrac{1}{2} \rho g H^2 \cdot w,
\]
per unit width $w$ of dam, acting at $H/3$ above the base. The
load grows quadratically with the water depth; a $100\,\si{\meter}
$ dam carries $\sim 10^7\,\text{N}/\text{m}$ of width, or $\sim
50\,\text{GN}$ across a $5\,\si{\kilo\meter}$ structure. The design
of the dam against this load is the subject of geotechnical and
structural engineering at scale; the failure mechanisms are
well-characterised:

\begin{itemize}[leftmargin=*]
  \item \engterm{Overtopping}. A flood event raises the
    upstream water level above the design head; water flows over
    the dam crest and erodes the downstream face (for an
    embankment dam) or scours the foundation (for a gravity
    dam). The discipline is the spillway: a designed overflow
    pathway that diverts excess water before the crest is
    overtopped. Most embankment-dam failures involve overtopping
    as the proximate mechanism.
  \item \engterm{Internal erosion (piping)}. Water under the
    dam's hydrostatic head migrates through the dam body or its
    foundation along preferential pathways. Fine soil particles
    are transported by the flow, enlarging the pathway, until a
    direct conduit through the dam forms. The dam then fails
    catastrophically by the upstream face collapsing into the
    pipe. The discipline is the seepage control: filter zones,
    drains, and grouting at the foundation interface.
  \item \engterm{Foundation failure}. The dam slides along its
    foundation under the horizontal component of the hydrostatic
    load, or rotates by toe-crushing under the vertical reaction.
    The discipline is the sliding-stability check (horizontal
    force times factor of safety against the available shear
    resistance) and the bearing-pressure check at the dam toe.
  \item \engterm{Structural failure of the dam body}. For a
    concrete gravity or arch dam, the dam's own mass and shape
    must carry the hydrostatic load as a structural member; for
    an embankment dam, the soil shear strength must resist the
    sliding force. Structural failure is the failure to satisfy
    the relevant code check, typically during an extreme
    loading event (an earthquake superposed on full reservoir).
\end{itemize}

The Malpasset Dam failure (Var, France, 2 December 1959), the
Vajont Dam landslide-induced overtopping (Italy, 9 October 1963),
and the Banqiao Dam failure (Henan, China, 8 August 1975, the
deadliest in history) are the canonical case studies in
dam-breach mechanics. Each has its own combination of design,
geological, and operational antecedents that are studied at
length in Volume X chapters 7 and 8 (forensic engineering of
large-scale failures); the failure mechanism's hydrostatic content
is, in each case, the quadratic-with-depth load against a
structure whose capacity was overestimated for the actual
foundation, geometry, or operating condition.

\subsection{Floating instability and capsize}

A ship, a floating platform, a barge, or any other floating body
has a stability calculation as part of its design. The
calculation is a function of the loading condition: the ballast,
the cargo distribution, the consumables (fuel, water), and the
weather (deck water, ice accretion). A ship that satisfies the
stability criterion in its design condition can fall below the
criterion in service if the loading condition departs from the
design.

The failure modes:

\begin{itemize}[leftmargin=*]
  \item \engterm{Loss of metacentric height by mass shift}. Cargo
    that shifts in the hold (a liquid that sloshes, a granular
    cargo that slides, a deck cargo that breaks free in heavy
    weather) raises the centre of mass $G$ and reduces
    $\overline{GM}$. A loss of $\overline{GM}$ that brings it
    near zero leaves the ship neutrally stable, vulnerable to
    capsize on the next wave-induced heel.
  \item \engterm{Loss of waterplane area by flooding}. A
    flooded compartment displaces fluid that no longer
    contributes to $I_{\text{wp}}$; the metacentric height
    decreases by a factor proportional to the lost waterplane
    area. The free surface inside the flooded compartment
    further reduces effective stability by the
    \engterm{free-surface effect}, an apparent reduction in
    $\overline{GM}$ equal to $\rho_f i/\rho V_{\text{disp}}$,
    with $i$ the second moment of the flooded compartment's
    waterplane area. Subdivision into watertight compartments
    limits both effects.
  \item \engterm{Ice accretion on deck}. In cold-weather
    operation, ice forms on the upper deck and superstructure,
    adding mass high in the hull. The mass shift raises $G$;
    the load further heels the ship and exposes more surface to
    spray and freezing. Several fishing-trawler capsizes in
    northern waters trace to this mechanism.
  \item \engterm{Heeling under turning}. A ship in a turn
    experiences a centripetal acceleration that produces an
    apparent body force outward; the apparent gravity tilts
    inward, and the ship heels outward in the turn. Combined
    with a marginal $\overline{GM}$, the heeling can exceed the
    angle of vanishing stability.
\end{itemize}

The diagnostic for floating stability is the
\engterm{stability booklet} maintained on every commercial
vessel, with the loading conditions for which the metacentric
height and the righting-arm curve have been computed, and the
restrictions on loading that the master must observe in service.
The stability booklet is a regulatory document under SOLAS
(International Convention for the Safety of Life at Sea); a
ship without an up-to-date stability booklet is not legally
permitted to operate.

\subsection{Surface-tension-driven leakage at small scales}

A third failure family sits at the small end of the length scale, where
the Bond number falls below one and surface tension competes with gravity.
A seal, a meniscus, or a wicking path that holds at large scale can leak at
small scale, and the same Young-Laplace pressure that lets a clean
capillary lift water lets a contaminated joint pass it. Two mechanisms
recur. The first is capillary wicking through a porous or microcracked
seal: a hairline crack of width $2r$ in a gasket sustains a capillary
pressure $2\sigma\cos\theta_c/r$ that, for a wettable crack a few
micrometres wide, runs to tens of kilopascals, enough to draw liquid
through a joint that shows no flow under the bulk hydrostatic head alone.
The second is the loss of a sealing meniscus when a surface changes from
non-wetting to wetting: a fuel or solvent that lowers the contact angle on
a previously dry seal turns capillary depression, which had kept liquid
out, into capillary rise, which pulls it in. The diagnostic is the Bond
number of the leak path, not of the vessel: a tank that is unambiguously
gravity-governed can still leak by a surface-tension-governed path through
a small enough flaw, and the inspection must look at the crack scale, not
the structure scale. The discipline is the same one the chapter installs
elsewhere, to check the Bond number on the length scale that actually
controls the physics before deciding which terms to keep.

\subsection{The diagnostic principle for fluid-statics failures}

Fluid-statics failure diagnostics start from the same
hydrostatic-force calculation that the design used. The question
is whether the actual loading condition matches the design
condition, whether the structure's actual capacity matches the
design capacity, and whether the supporting systems (spillways,
drains, watertight subdivision) are functioning as designed. A
fluid-statics design that has been completed correctly is
durable; failures usually trace to a departure between the design
condition and the service condition, not to an error in the
hydrostatic mathematics. The mathematical content of the chapter
is durable; the engineering discipline is the discipline of
keeping the service condition within the design envelope.

\begin{principle}[The hydrostatic load grows quadratically with
depth]
The total hydrostatic force on a vertical retaining surface of
depth $H$ is $F = \tfrac{1}{2}\rho g H^2$ per unit width. The
quadratic dependence has two consequences: doubling the depth of
water multiplies the wall load by four; the line of action sits
at $H/3$ above the base, so the overturning moment grows as $H^3$.
Dams, basement walls, tank walls, and hull plates are all
governed by the same scaling, and every increase in
design head must be checked against this growth.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Twenty-four-hour barometric record]
The reader records atmospheric pressure continuously for at least
$24\,\text{hours}$ at a fixed location, correlates the record
with weather observations, and produces a calibrated pressure
record with quantified uncertainty. The deliverable is a
twenty- to thirty-page engineering report.

\textbf{Track}. Instrumentation and field measurement.
\textbf{Hazard class}: none.

\textbf{Instrument}. The reader uses a barometric-pressure
sensor: a phone with a built-in MEMS pressure sensor (current
phones contain BMP-class sensors with $\pm 1\,\text{hPa}$
typical accuracy), a USB-connected sensor module (BMP280 or
BME280, $\pm 0.12\,\text{hPa}$ relative accuracy), a household
weather station, or a Bourdon-tube barometer with a hand-read
dial. The choice of instrument is documented with its specified
accuracy.

\textbf{Calibration}. The reader either calibrates the
instrument against an independent reference (a co-located
weather station, an online METAR report from a nearby airport)
or records the instrument's specified accuracy as a published
uncertainty.

\textbf{Procedure}. The reader records the pressure at a fixed
location for at least $24\,\text{hours}$, with a sampling
interval of $10\,\text{minutes}$ or shorter. The reader records
qualitative weather observations every $3\,\text{hours}$:
cloud cover, precipitation, wind direction.

\textbf{Analysis}. The reader produces:

\begin{itemize}[leftmargin=*]
  \item A plot of pressure versus time over the recording
    interval, with the measurement uncertainty shown.
  \item Identification of any pressure trends: rising, falling,
    diurnal cycle (the semi-diurnal atmospheric tide produces a
    typical $\pm 1.5\,\text{hPa}$ peak-to-peak variation in mid-
    latitudes, with maxima near $10$ and $22$ local solar time).
  \item Correlation with weather observations: a falling
    pressure typically precedes a cold front; a rising pressure
    typically follows a frontal passage.
  \item Comparison with the official meteorological record for
    the same period (METAR, regional weather service), if
    available.
  \item An uncertainty budget: instrument accuracy, calibration
    drift, location-of-instrument effects (indoor versus
    outdoor, ventilation effects on the sensor).
\end{itemize}

\textbf{Reflection ($1{,}000$ to $2{,}000$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
  \item Was the diurnal cycle visible? If not, why not?
  \item What pressure change accompanied the most weather-
    relevant event of the recording interval?
  \item What is the smallest pressure change the instrument can
    reliably resolve, given the noise floor of the record?
  \item What is the implied altimeter sensitivity (pressure
    change per metre of altitude) at the location's typical
    pressure?
\end{itemize}

\textbf{Deliverable}. The pressure record (plotted and
tabulated), the weather log, the uncertainty budget, the
correlation analysis, and the reflection.

\textbf{Time}. The recording is unattended; the analysis and
write-up require four to eight hours.
\end{project}

\section{Exercises}

The exercises mix calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A swimming pool is $25\,\si{\meter}$ long, $10\,\si{\meter}$
wide, and $2\,\si{\meter}$ deep at the deep end. Compute the
gauge pressure at the bottom of the deep end and the total
hydrostatic force on the deep-end wall (assuming the deep-end
wall is vertical and $10\,\si{\meter}$ wide). State the location
of the resultant force above the floor.

\paragraph{Solution sketch.} Gauge pressure at the bottom:
$p = \rho g h = 1000 \times 9.81 \times 2 = 19.62\,\text{kPa}$. The
deep-end wall is a vertical rectangle of depth $H = 2\,\si{\meter}$ and
width $w = 10\,\si{\meter}$ with the surface at its top, so the force is
the pressure-prism area, $F = \tfrac12 \rho g H^2 w = \tfrac12 \times
1000 \times 9.81 \times 4 \times 10 \approx 196\,\text{kN}$. The
resultant acts at the centroid of the triangular prism, $H/3 =
0.67\,\si{\meter}$ above the floor.
\end{exercise}

\begin{exercise}
A U-tube manometer is partially filled with mercury and connected
at one leg to a gas line. The mercury level in the open leg is
$120\,\si{\milli\meter}$ higher than the mercury level in the
gas-connected leg. Compute the gauge pressure of the gas.

\paragraph{Solution sketch.} The open leg standing higher means the gas
pushes the mercury down on its side, so the gas is above atmospheric.
Neglecting the light gas column, $p_{\text{gauge}} = \rho_{\text{Hg}} g
\Delta h = 13\,600 \times 9.81 \times 0.120 \approx 16.0\,\text{kPa}$.
That is about $0.16\,\text{atm}$, or $120\,\text{mmHg}$ by the unit's
own definition. Carrying the gas column explicitly changes the answer by
under $0.01\,\%$ (see \texttt{code/manometer.py}).
\end{exercise}

\begin{exercise}
A submerged hatch on a ship is rectangular, $0.8\,\si{\meter}$
wide and $1.2\,\si{\meter}$ tall, with its top edge at depth
$3\,\si{\meter}$ below the water surface. The hatch is vertical.
Compute the resultant hydrostatic force on the hatch and the
depth of the centre of pressure.

\paragraph{Solution sketch.} Centroid depth $h_C = 3 + 1.2/2 =
3.6\,\si{\meter}$. Area $A = 0.8 \times 1.2 = 0.96\,\text{m}^2$.
Force $F = \rho g h_C A = 1000 \times 9.81 \times 3.6 \times 0.96
\approx 33.9\,\text{kN}$. Second moment about the centroidal horizontal
axis $I_{xx,C} = w b^3/12 = 0.8 \times 1.2^3 / 12 = 0.1152\,\text{m}^4$.
Centre of pressure $h_P = h_C + I_{xx,C}/(h_C A) = 3.6 + 0.1152/(3.6
\times 0.96) \approx 3.6 + 0.033 \approx 3.63\,\si{\meter}$. The offset
is small because the hatch is small and deep, so the pressure is nearly
uniform across it.
\end{exercise}

\begin{exercise}
A spherical buoy of radius $0.25\,\si{\meter}$ and mass $20\,
\si{\kilogram}$ floats in seawater ($\rho = 1025\,\si{\kilogram
\per\meter\cubed}$). Compute the depth of the buoy's centre
below the water surface in static equilibrium.

\paragraph{Solution sketch.} Floating equilibrium: buoyancy equals
weight, $\rho_f g V_{\text{disp}} = m g$, so the submerged volume is
$V_{\text{disp}} = m/\rho_f = 20/1025 \approx 0.01951\,\text{m}^3$. The
full sphere has volume $\tfrac43\pi r^3 = \tfrac43\pi (0.25)^3 \approx
0.06545\,\text{m}^3$, so the buoy floats with about $30\,\%$ submerged.
The submerged volume of a sphere as a function of cap depth $d$ (from
the bottom) is $V = \pi d^2 (3r - d)/3$. Setting this equal to
$0.01951\,\text{m}^3$ with $r = 0.25$ and solving the cubic numerically
gives $d \approx 0.166\,\si{\meter}$. The centre of the sphere sits at
$d - r = 0.166 - 0.25 \approx -0.084\,\si{\meter}$ relative to the
surface, that is, about $0.084\,\si{\meter}$ above the waterline (the
centre rides above water because less than half the buoy is submerged).
\end{exercise}

\begin{exercise}
A wooden cube of side $0.3\,\si{\meter}$ and density $600\,\si{
\kilogram\per\meter\cubed}$ floats in fresh water. Compute the
depth to which the cube sinks below the water surface.

\paragraph{Solution sketch.} A cube floats upright with submerged
fraction equal to the density ratio: $T/L = \rho_{\text{body}}/\rho_f =
600/1000 = 0.6$. So the cube sinks to a depth $T = 0.6 \times 0.3 =
0.18\,\si{\meter}$, leaving $0.12\,\si{\meter}$ of freeboard. (The cube
flat-floating in this attitude is the simplest buoyancy balance:
displaced volume $L^2 T$ times $\rho_f$ equals body volume $L^3$ times
$\rho_{\text{body}}$, and the side $L^2$ cancels.)
\end{exercise}

\begin{exercise}
A vertical dam, $30\,\si{\meter}$ tall and $200\,\si{\meter}$
long, retains a reservoir whose water surface reaches the top of
the dam. Compute the total horizontal hydrostatic force on the
dam and the height above the dam base at which the resultant
acts.

\paragraph{Solution sketch.} Per unit width, $F/L = \tfrac12 \rho g H^2
= \tfrac12 \times 1000 \times 9.81 \times 30^2 \approx 4.41\,\text{MN/m}$.
Total over $L = 200\,\si{\meter}$: $F \approx 882\,\text{MN}$. The
resultant acts at the centroid of the triangular pressure prism, at
$H/3 = 10\,\si{\meter}$ above the base. This matches the worked vertical
dam of Section 10.7; load grows as $H^2$ and the overturning moment as
$H^3$.
\end{exercise}

\begin{exercise}
A cylindrical tank of inner radius $1\,\si{\meter}$ and depth
$3\,\si{\meter}$ is filled with water. The tank rotates about
its vertical axis at $\omega = 5\,\text{rad/s}$. Compute the
height difference between the water surface at the centre and
at the wall, and the gauge pressure at the bottom at the centre
and at the wall.

\paragraph{Solution sketch.} The free surface is a paraboloid,
$z(r) = z_0 + \omega^2 r^2/(2g)$. The rise from centre to wall is
$\Delta z = \omega^2 R^2/(2g) = 25 \times 1/(2 \times 9.81) \approx
1.27\,\si{\meter}$ (assuming the tank is deep enough that the water does
not spill or expose the bottom; here depth $3\,\si{\meter}$ comfortably
holds it). Pressure at the bottom, $p(r,z=0) = \rho g d_{\text{local}}$
where $d_{\text{local}}$ is the local surface-to-bottom depth. If the
mean depth stays $3\,\si{\meter}$ (volume conserved), the centre surface
drops by $\Delta z/2 \approx 0.64\,\si{\meter}$ and the wall surface
rises by the same, so bottom pressures are: centre $\approx \rho g (3 -
0.64) \approx 23.2\,\text{kPa}$; wall $\approx \rho g (3 + 0.64) \approx
35.7\,\text{kPa}$. The difference, $12.5\,\text{kPa}$, equals
$\tfrac12\rho\omega^2 R^2 = \tfrac12 \times 1000 \times 25 \times 1$,
the centrifugal pressure term.
\end{exercise}

\begin{exercise}
A water tank truck is decelerating at $3\,\text{m/s}^2$. The tank
is rectangular in cross-section and is partially full. Compute
the angle the water surface makes with the horizontal.

\paragraph{Solution sketch.} The free surface is normal to the
effective gravity $\vec{g}_{\text{eff}} = -g\hat{k} - a\hat{i}$, so it
tilts from horizontal by $\alpha = \tan^{-1}(a/g) = \tan^{-1}(3/9.81)
\approx 17.0^{\circ}$. The surface rises toward the rear of the tank
(the deceleration throws the water forward, but the surface tips so its
high end is at the rear in the frame of the truck; the equivalent
statement is that the surface tilts down in the direction of
deceleration).
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the hydrostatic equation $\nabla p = \rho \vec{g}$ from
force balance on an infinitesimal fluid element. State the two
assumptions about the fluid's stress state that the derivation
uses.

\paragraph{Solution sketch.} Take a box $dx\,dy\,dz$. The net pressure
force in $x$ is $[p(x) - p(x+dx)]\,dy\,dz = -(\partial p/\partial x)\,
dV$, and likewise in $y$ and $z$. Add the body force $\rho\vec{g}\,dV$
and set the sum to zero for static equilibrium, then divide by $dV$:
$-\nabla p + \rho\vec{g} = 0$, that is $\nabla p = \rho\vec{g}$. The two
stress-state assumptions are (i) the fluid sustains no shear stress at
rest, so the only surface stress is the normal pressure, and (ii)
pressure is isotropic (a scalar), the same on a face of any orientation,
which is the tetrahedron result of Section 10.1.
\end{exercise}

\begin{exercise}
Derive the formula $F = p_C A$ for the resultant hydrostatic
force on a flat submerged surface of arbitrary shape, with $p_C$
the pressure at the centroid and $A$ the area.

\paragraph{Solution sketch.} The pressure on an area element at depth
$h$ is $p = \rho g h$, so $F = \int_A \rho g h\,dA = \rho g \int_A h\,dA$.
By the definition of the centroid, $\int_A h\,dA = h_C A$, where $h_C$ is
the centroidal depth. Hence $F = \rho g h_C A = p_C A$ with $p_C =
\rho g h_C$. The result holds for any plate shape and any inclination
because the only fact used is the linear pressure-depth law and the
centroid definition.
\end{exercise}

\begin{exercise}
Derive the formula for the centre of pressure $h_P = h_C +
I_{xx,C}/(h_C A)$ for a vertical flat submerged surface, with
$I_{xx,C}$ the second moment of area about the horizontal
centroidal axis.

\paragraph{Solution sketch.} The centre of pressure is the depth where
the resultant produces the same moment about the free-surface axis as
the distributed load. Moment of the pressure about the surface axis:
$M = \int_A h\,p\,dA = \rho g \int_A h^2\,dA = \rho g\, I_{xx,O}$, where
$I_{xx,O}$ is the second moment about the surface axis. The resultant
$F = \rho g h_C A$ acting at $h_P$ gives the same moment: $h_P F =
\rho g I_{xx,O}$, so $h_P = I_{xx,O}/(h_C A)$. Apply the parallel-axis
theorem $I_{xx,O} = I_{xx,C} + h_C^2 A$ to get $h_P = h_C +
I_{xx,C}/(h_C A)$. The correction $I_{xx,C}/(h_C A)$ is always positive,
so the centre of pressure lies below the centroid.
\end{exercise}

\begin{exercise}
Derive Archimedes' principle, $F_B = \rho_f g V_{\text{disp}}$,
by integrating the hydrostatic pressure over the surface of an
arbitrary submerged body.

\paragraph{Solution sketch.} The net pressure force is the closed
surface integral $\vec{F} = -\oint_S p\,\hat{n}\,dS$ (pressure pushes
inward along the outward normal $\hat{n}$). Write $p = p_0 + \rho g z'$
with $z'$ depth below the surface, and use the divergence theorem,
$\oint_S p\,\hat{n}\,dS = \int_V \nabla p\,dV = \int_V \rho g \hat{k}_{
\downarrow}\,dV$. The constant part $p_0$ integrates to zero over a
closed surface; the depth-dependent part gives a net upward force equal
to $\rho g V_{\text{disp}}$. So $F_B = \rho_f g V_{\text{disp}}$,
directed upward, acting through the centroid of the displaced volume.
\end{exercise}

\begin{exercise}
Derive the metacentric-height formula $\overline{GM} = I_{\text{
wp}}/V_{\text{disp}} - \overline{BG}$ for a floating body
perturbed by a small angle from equilibrium. State the
small-angle assumption.

\paragraph{Solution sketch.} Heel by small $\theta$. A wedge of fluid
leaves the rising side and an equal wedge enters the descending side;
the transferred volume shifts the centre of buoyancy sideways by
$\overline{BB'} = \theta I_{\text{wp}}/V_{\text{disp}}$, where
$I_{\text{wp}} = \int y^2\,dA$ is the waterplane second moment about the
roll axis (the wedge derivation is in the mastery box of Section 10.5).
The metacentre $M$ is the point where the new buoyancy line meets the
centreline, at height $\overline{BM} = \overline{BB'}/\theta =
I_{\text{wp}}/V_{\text{disp}}$ above $B$. Then $\overline{GM} =
\overline{BM} - \overline{BG} = I_{\text{wp}}/V_{\text{disp}} -
\overline{BG}$. The small-angle assumption is that the waterplane shape
is unchanged to first order in $\theta$, so $\tan\theta \approx \theta$
and the wedge volume is linear in $\theta$; it fails at large heel where
the deck edge immerses or the bilge emerges.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the depth at which the gauge pressure equals atmospheric
pressure in fresh water. Comment on the practical significance.

\paragraph{Reference estimate.} Set $\rho g h = p_{\text{atm}}$:
$h = 101\,325/(1000 \times 9.81) \approx 10.3\,\si{\meter}$. So at about
ten metres' depth the gauge pressure equals one atmosphere and the
absolute pressure has doubled to two atmospheres. The practical
consequence: a breath-hold diver's lungs halve in volume by ten metres;
scuba gas consumption doubles; and a suction lift of water cannot exceed
this depth because the atmosphere can push only one atmosphere's worth
of column.
\end{exercise}

\begin{exercise}
Estimate the maximum depth a free-diving human can reach without
external assistance, given physiological constraints on lung
compression and blood gas exchange. State assumed limits and
identify which physiological factor most directly limits the
depth.

\paragraph{Reference estimate.} A naive limit once taught was lung
compression: total lung capacity $\sim 6\,\text{L}$, residual volume
$\sim 1.5\,\text{L}$, so by Boyle's law full compression to residual
volume occurs near a fourfold pressure increase, $p/p_0 \approx 4$, that
is depth $\approx 3 \times 10.3 \approx 30\,\si{\meter}$. Observed
no-limits free-dive records exceed $200\,\si{\meter}$, so the lung-volume
limit is not the operative one: blood and tissue shift into the thorax
(blood shift) and partly fill the compressing lungs, so the rib cage is
not crushed. The actual limiting factors are gas exchange and
consciousness: oxygen partial pressure, nitrogen narcosis, and the
hypoxic blackout on ascent when falling pressure drops alveolar oxygen
partial pressure below the level that sustains consciousness. The
most direct limiter is therefore blood-gas and central-nervous-system
oxygen supply, not mechanical lung compression.
\end{exercise}

\begin{exercise}
Estimate the volume of water displaced by a typical passenger car
($\sim 1500\,\si{\kilogram}$) if it falls into a lake. State
whether the car floats or sinks (assume sealed doors and windows
initially) and the implication for emergency egress.

\paragraph{Reference estimate.} A passenger car's bounding box is about
$4.5 \times 1.8 \times 1.5 \approx 12\,\text{m}^3$, but the watertight
sealed volume (cabin plus some structure) is perhaps $3\text{ to }4\,
\text{m}^3$. To float, the car must displace $1500/1000 = 1.5\,\text{m}^3$
of water. Since the sealed volume exceeds $1.5\,\text{m}^3$ at first, a
car with intact seals floats briefly, low in the water and usually
nose-down (the engine mass is forward). It then sinks as water enters
through door seals, vents, and the cabin. The egress implication: the
window of time when the car still floats is the window to open a window
or door, because once the cabin floods the external hydrostatic load on
a submerged door (pressure times door area) exceeds human strength.
Break the window early; do not wait for pressure equalisation, which
arrives only after the cabin is full.
\end{exercise}

\begin{exercise}
Estimate the pressure at the deepest point of the Mariana Trench
(approximately $11\,\si{\kilo\meter}$ depth), including the
compressibility correction for seawater. Compare to the
incompressible-water estimate.

\paragraph{Reference estimate.} Incompressible estimate: $p = \rho g h
= 1025 \times 9.81 \times 11\,000 \approx 1.106 \times 10^8\,\text{Pa}
\approx 1090\,\text{atm}$. Seawater compresses under this load, raising
the density with depth; with a bulk modulus $K \approx 2.3\,\text{GPa}$,
density at the bottom rises by roughly $\Delta\rho/\rho \approx p/K
\approx 0.11/2.3 \approx 4.8\,\%$. Integrating $dp = \rho(p) g\,dz$ with
density increasing toward the bottom adds about $2\text{ to }3\,\%$ to
the pressure, giving $\approx 1.13 \times 10^8\,\text{Pa} \approx
1120\,\text{atm}$. Measured values at Challenger Deep are about
$1.16 \times 10^8\,\text{Pa}$; the residual gap is the temperature and
salinity profile, which the simple bulk-modulus correction omits. The
incompressible estimate is within a few per cent, which is the usual
verdict: compressibility matters at the level of engineering margin, not
at the level of order of magnitude.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Numerically integrate the hydrostatic equation in the standard
atmosphere from sea level to $32\,\si{\kilo\meter}$. Compare the
computed pressure-altitude profile to the published ISA tables.

\paragraph{Solution sketch.} Integrate $dp/dz = -pMg/(RT(z))$ layer by
layer with the piecewise-linear ISA temperature profile (lapse rate
$6.5\,$K/km to $11\,$km, isothermal to $20\,$km, then rising). Each
constant-lapse layer integrates in closed form, $p = p_b (T/T_b)^{Mg/(RL)}$,
or use a small-step RK4 on the ODE; both should reproduce
$\approx 22.6\,\text{kPa}$ at the tropopause and $\approx 0.87\,\text{kPa}$
at $32\,\si{\kilo\meter}$. The reference implementation is
\texttt{code/standard\_atmosphere.py} and its output table is
\texttt{data/isa\_profile.csv}; agreement with published ISA values is to
better than $0.1\,\%$. The single-exponential isothermal model
overshoots aloft, the contrast plotted in
Figure~\ref{fig:vol03ch10:atmosphere}.
\end{exercise}

\begin{exercise}
Implement a numerical solver for the equilibrium attitude of a
floating rectangular box, given the box dimensions, mass
distribution, and water density. Verify that the solver returns
a stable solution for a beamy hull and an unstable solution for
a tall narrow hull.

\paragraph{Solution sketch.} For an upright homogeneous box the draft is
$T = (\rho_b/\rho_f)D$; floating fails if $T \ge D$. Compute
$\overline{KB} = T/2$, $\overline{KG} = D/2$, $\overline{BM} =
I_{\text{wp}}/V_{\text{disp}} = (LB^3/12)/(LBT) = B^2/(12T)$, and
$\overline{GM} = \overline{KB} + \overline{BM} - \overline{KG}$. The
solver returns stable when $\overline{GM} > 0$. The reference is
\texttt{code/metacentre.py}; on the test cases in
\texttt{data/box\_stability.csv} the beamy barge ($B = 4$) gives
$\overline{GM} \approx +1.58\,\si{\meter}$ (stable) and the tall narrow
hull ($B = 1$) gives $\overline{GM} \approx -0.92\,\si{\meter}$
(unstable), confirming the solver. A fuller solver that sweeps heel
angle and finds the equilibrium where the righting moment vanishes
recovers the same small-angle verdict and additionally locates the angle
of vanishing stability.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a rectangular swimming-pool wall, $2\,\si{\meter}$ deep
and $25\,\si{\meter}$ long, in reinforced concrete. Specify the
wall thickness and the reinforcement based on the hydrostatic
load. State the assumed concrete and steel properties.

\paragraph{Model-answer sketch.} Loading: triangular pressure prism,
$F/L = \tfrac12\rho g H^2 = \tfrac12 \times 1000 \times 9.81 \times 4
\approx 19.6\,\text{kN/m}$, acting at $H/3 = 0.67\,\si{\meter}$ above the
base. Treat the wall as a vertical cantilever fixed at the floor slab.
Bending moment at the base per metre of length: $M = (F/L) \times H/3
\approx 19.6 \times 0.67 \approx 13.1\,\text{kN}\cdot\text{m/m}$. With
concrete $f'_c = 30\,\text{MPa}$ and reinforcing steel $f_y =
500\,\text{MPa}$, a wall thickness of $200\text{ to }250\,\si{
\milli\meter}$ gives ample lever arm; required steel area from
$M = A_s f_y\, (0.9d)$ with $d \approx 0.2\,\si{\meter}$ is
$A_s \approx 13\,100/(500\times10^6 \times 0.18) \approx 1.5 \times
10^{-4}\,\text{m}^2/\text{m} = 145\,\text{mm}^2/\text{m}$, easily met by
minimum reinforcement (for example $12\,\si{\milli\meter}$ bars at
$300\,\si{\milli\meter}$ on the water face). The design is governed by
minimum reinforcement and crack control (watertightness), not by the
small hydrostatic moment; state the assumed cover, exposure class, and a
load factor of about $1.5$ on the water pressure.
\end{exercise}

\begin{exercise}
Design a flat circular hatch cover for a submersible vessel
designed to operate at $100\,\si{\meter}$ depth. The hatch is
$0.6\,\si{\meter}$ in diameter and is mounted in a horizontal
deck. Specify the hatch thickness and the seal-pre-compression
requirement, using a steel of yield strength $400\,\text{MPa}$
with a safety factor of $3$.

\paragraph{Model-answer sketch.} External pressure at $100\,\si{\meter}$
seawater: $p = \rho g h = 1025 \times 9.81 \times 100 \approx 1.01\,
\text{MPa} \approx 10\,\text{atm}$. Treat the hatch as a clamped
circular plate of radius $a = 0.3\,\si{\meter}$ under uniform pressure.
Maximum bending stress (clamped edge) $\sigma_{\max} = 0.75\, p\,
(a/t)^2$. Set $\sigma_{\max} = f_y/\text{SF} = 400/3 \approx 133\,\text{
MPa}$ and solve for $t$: $(a/t)^2 = 133\times10^6/(0.75 \times 1.01
\times10^6) \approx 176$, so $a/t \approx 13.3$ and $t \approx
0.3/13.3 \approx 23\,\si{\milli\meter}$. Round up to $25\,\si{
\milli\meter}$. Seal pre-compression: the seal must stay compressed at
the surface (zero external pressure) yet not over-compress at depth, so
specify a gasket pre-load that exceeds the bolt-relaxation and
thermal-contraction allowance at the surface; at depth the external
pressure adds to the seating load, which helps the seal rather than
unseats it (the door is pushed shut). State the plate-theory assumption
(thin plate, $t/a < 0.1$, here $25/300 \approx 0.08$, satisfied).
\end{exercise}

\begin{exercise}
Design a rectangular barge to carry a $20\,\si{\tonne}$ load
in fresh water with at least $0.3\,\si{\meter}$ of freeboard
above the loaded waterline. Specify the barge dimensions
(length, beam, depth) and verify the metacentric height in the
loaded condition.

\paragraph{Model-answer sketch.} Take a barge $L = 8\,\si{\meter}$,
$B = 3\,\si{\meter}$, depth $D = 1.2\,\si{\meter}$, hull mass say
$4\,\si{\tonne}$, so total loaded mass $24\,\si{\tonne}$. Required draft:
$T = m/(\rho_f LB) = 24\,000/(1000 \times 8 \times 3) = 1.0\,\si{\meter}$,
leaving freeboard $D - T = 0.2\,\si{\meter}$, which is below the
$0.3\,\si{\meter}$ requirement, so increase $D$ to $1.3\,\si{\meter}$ for
$0.3\,\si{\meter}$ freeboard. Stability: $\overline{KB} = T/2 = 0.5$;
$\overline{BM} = B^2/(12T) = 9/12 = 0.75\,\si{\meter}$; the loaded
centre of mass depends on where the cargo sits, but for cargo near the
deck take $\overline{KG} \approx 0.9\,\si{\meter}$. Then
$\overline{GM} = 0.5 + 0.75 - 0.9 = 0.35\,\si{\meter} > 0$, stable, with
a modest but acceptable metacentric height. If the cargo rides higher,
$\overline{KG}$ rises and $\overline{GM}$ can go negative, so the design
must constrain the vertical centre of gravity of the load, which is the
loading-plan discipline of Section 10.7.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A residential basement wall has bowed inward by approximately
$15\,\si{\milli\meter}$ at midheight. The bowing has appeared
during a season of unusually heavy rainfall. Identify the
loading mechanism, the failure mode at risk, and the corrective
action.

\paragraph{Solution sketch.} Loading mechanism: a rising water table
during heavy rain raises the lateral hydrostatic (and saturated-soil)
pressure on the outside of the wall; the load grows quadratically with
the saturated depth, and the wall bends inward under it. Failure mode at
risk: flexural cracking and progressive inward collapse of an
inadequately reinforced or unreinforced masonry/concrete wall, the
classic basement-wall bowing failure; a $15\,\si{\milli\meter}$ midheight
bow is a warning that the wall is near its bending capacity. Corrective
action: relieve the load (improve site drainage, install or clear
perimeter footing drains, regrade to shed surface water away from the
wall) and reinforce the wall (carbon-fibre straps, steel I-beam braces,
or wall anchors tied to the soil outside). The discipline is
load-then-capacity: reduce the hydrostatic driving load first, then
restore the structural capacity.
\end{exercise}

\begin{exercise}
A water-storage tank designed for a maximum head of $5\,\si{\meter}
$ has developed a circumferential crack at the base of its
cylindrical wall. The crack is on the inside face and is open
under static head. Identify the structural failure mode and the
inspection that should have detected the precursor.

\paragraph{Solution sketch.} A circumferential crack at the base of a
cylindrical tank wall, open on the inside face under head, is a
vertical-bending failure: the wall acts as a vertical cantilever
restrained by the floor, and the hydrostatic pressure (maximum at the
base, $p = \rho g H = 1000\times9.81\times5 \approx 49\,\text{kPa}$)
bends the wall outward at the base, putting the inside face into tension
near the floor-wall junction. The hoop tension that the head also
produces is separate; the circumferential horizontal crack at the base
points to the vertical-bending (or restrained-base) mode, often worsened
by differential settlement or a stiff base detail. The precursor
inspection: periodic visual and crack-width survey of the inside base
junction during a full-head condition, plus settlement monitoring;
hairline cracking on the inside face at the base is the early indicator,
and a strain-gauge or crack-gauge at the junction would have flagged the
growth before the crack opened under static head.
\end{exercise}

\begin{exercise}
A fishing vessel has developed a permanent list of $5^{\circ}$
to one side. The list does not correct after fuel transfer
between port and starboard tanks. Identify the most probable
causes (in order of likelihood) and the inspection that
discriminates between them.

\paragraph{Solution sketch.} A permanent list that does not correct with
symmetric fuel transfer means the heeling moment is not in the
transferable fuel. Most probable causes, in order: (1) asymmetric
flooding or water ingress in a low compartment or void on the list side
(bilge water, a leak, blocked drainage), adding off-centre weight; (2) a
fixed off-centre weight, such as a shifted or wrongly stowed catch, gear,
or ballast; (3) a transverse shift of the centre of gravity from
asymmetric ice or deck load; (4) less likely, an asymmetric hull
deformation or trapped water in a fish hold. The discriminating
inspection: sound and inspect all voids, bilges, and tanks on the list
side for water (a tank-sounding survey), which separates flooding
(case 1) from a solid off-centre weight (cases 2 to 3); if soundings are
dry, conduct a weight and stowage survey. The free-surface effect of any
trapped water also reduces $\overline{GM}$, so finding and removing it
addresses both the list and a stability deficit.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A concrete gravity dam fails by sliding along its foundation
during an extreme flood event in which the spillway capacity is
exceeded and the reservoir rises $2\,\si{\meter}$ above the
design crest. Argue, in $200$ to $300$ words, the contribution
of the additional hydrostatic head to the sliding load and
identify which design discipline should have anticipated this
combination of events.

\paragraph{Solution sketch.} The horizontal hydrostatic thrust grows as
$\tfrac12\rho g H^2$, so raising the head from the design crest $H$ to
$H + 2$ multiplies the sliding load by $((H+2)/H)^2$; on a $30\,\si{
\meter}$ dam, $((32)/30)^2 \approx 1.14$, a $14\,\%$ increase in
driving force. The argument should add that the rising reservoir also
raises the uplift (pore) pressure under the base, which reduces the
effective normal force and thus the available frictional shear
resistance, so the factor of safety against sliding falls faster than
the $14\,\%$ load increase alone suggests; overtopping erosion of the
downstream toe further degrades resistance. The discipline that should
have anticipated the combination is hydrology and spillway design feeding
the geotechnical sliding-stability check: the extreme-flood load case
(probable maximum flood with reduced spillway capacity, combined with
full uplift) is a required load combination, and treating the design
crest as the maximum head omitted the very event that occurred. Name the
quadratic load growth, the uplift coupling, and the missing flood load
case.
\end{exercise}

\begin{exercise}
A cargo ship capsizes in moderate weather when the cargo (steel
coils) breaks free in the hold and shifts to one side. Argue, in
$200$ to $300$ words, the mechanism by which the cargo shift
reduced the metacentric height and identify the loading
discipline that should have prevented the unsecured stow.

\paragraph{Solution sketch.} A transverse cargo shift moves the centre
of gravity $G$ off the centreline and upward, creating a fixed heeling
moment $w \times d$ ($w$ the shifted weight, $d$ its transverse travel).
The ship heels until the righting moment balances this offset; because
the heel reduces the effective righting arm $\overline{GZ}$ and the
shifted mass also raises $\overline{KG}$ (cutting $\overline{GM} =
\overline{KB} + \overline{BM} - \overline{KG}$), each successive wave can
drive a larger heel with less restoring moment. Once the steady heel
plus a wave-induced roll exceeds the angle of vanishing stability, the
righting arm goes negative and the ship capsizes. The argument should
distinguish the static heeling offset (a fixed moment) from the loss of
reserve stability (reduced $\overline{GM}$ and $\overline{GZ}$ curve):
the coils did not need to move far to be lethal once $\overline{GM}$ was
already marginal. The loading discipline that should have prevented it is
cargo securing and stowage planning under the vessel's approved cargo
securing manual: heavy unit cargo such as steel coils must be lashed and
chocked to a calculated standard so it cannot shift in the rolling
expected for the voyage, and the stability booklet's loaded condition
assumes that securing holds.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer asks whether gauge pressure or absolute
pressure should be used in a structural stress calculation on a
pressure-vessel wall. Write a one-paragraph reply that takes the
question seriously and identifies the regimes in which each
choice is correct.

\paragraph{Model-answer sketch.} The wall stress responds to the
pressure difference across the wall, so for a vessel surrounded by
atmosphere the load is the gauge pressure (internal absolute minus
external atmospheric), and using gauge pressure for the hoop and
longitudinal stress is correct. Absolute pressure enters where the
external reference is not atmospheric: a vessel in a vacuum or in space
carries the full internal absolute pressure as the differential; a
submerged vessel carries internal absolute minus external water
pressure, which can drive the wall into compression (an implosion risk)
rather than tension. Absolute pressure is also the right quantity inside
any equation of state (the gas law that sets the internal pressure from
temperature and mass) and in compressible-flow work. The discipline is
to write the actual pressure difference across the wall, naming both
sides, rather than reaching for gauge or absolute by habit; record the
reference explicitly in the calculation so a reviewer can check it.
\end{exercise}

\begin{exercise}
A senior colleague argues that surface tension is a curiosity
of physics class and never matters in real engineering. Write a
one-paragraph reply that identifies three engineering contexts
in which surface tension is load-bearing.

\paragraph{Model-answer sketch.} Three contexts where surface tension
carries real engineering weight: (1) inkjet and other drop-on-demand
printing, where the Young-Laplace pressure $2\sigma/R$ sets the
ejection and the break-off of microscopic droplets, and the whole device
fails if the surface tension of the ink is out of specification; (2)
microfluidic and lab-on-chip devices, and capillary action in wicks,
heat-pipe wicks, and soil moisture, where at small length scales the
Bond number is well below one and surface tension, not gravity, moves
the fluid; (3) the alveolar surfactant in the lung and the analogous
problem of bubble and foam stability in chemical reactors and food
processing, where the Young-Laplace pressure across curved interfaces
governs whether small bubbles collapse into large ones. The closing
point is the Bond number $\text{Bo} = \rho g L^2/\sigma$: surface
tension is negligible for dams and ship hulls precisely because their
$L$ is large, and load-bearing for droplets and capillaries precisely
because their $L$ is small. The colleague is right about big structures
and wrong about everything below the capillary length, about
$2.7\,\si{\milli\meter}$ for water.
\end{exercise}

\begin{exercise}
Identify a static-fluid load in your immediate environment whose
magnitude you have never explicitly calculated (a water heater's
tank pressure, the load on a refrigerator-door seal, the
hydrostatic head on a fish tank's wall, the suction on a bath
plug). State, in $200$ to $300$ words, the loading geometry, the
resultant force, the location of its line of action, and the
failure mode you would inspect for first.

\paragraph{Model-answer sketch.} A complete answer picks one object and
carries the four parts. Worked illustration, a $200\,\si{\liter}$ home
aquarium, glass front $1.2\,\si{\meter}$ wide and $0.6\,\si{\meter}$
deep, water to the top. Loading geometry: a triangular pressure prism on
the vertical front pane, zero at the surface to $\rho g H = 1000 \times
9.81 \times 0.6 \approx 5.9\,\text{kPa}$ at the base. Resultant per pane:
$F = \tfrac12\rho g H^2 w = \tfrac12 \times 1000 \times 9.81 \times 0.36
\times 1.2 \approx 2.1\,\text{kN}$, about the weight of two adults. Line
of action: $H/3 = 0.2\,\si{\meter}$ above the base. Failure mode to
inspect first: the silicone seam joints at the bottom corners, which
carry the peeling and shear load where the pressure is highest, and the
glass-pane bending stress at midspan of the front. The point of the
exercise is to convert an everyday object into the same four-step
calculation the chapter applies to dams: geometry, resultant, line of
action, failure mode.
\end{exercise}
