\chapter{Optics}
\label{vol04:ch11}

\epigraph{An optical system is a means for transforming wavefronts,
and the operations it performs are limited only by the imagination
of the designer and the discipline of the manufacturer.}{Eugene
Hecht, \emph{Optics}, 5th ed.\ (Pearson, 2017), paraphrased from
Chapter~5 \cite[Chapter~5]{text:hecht-optics}.}

\chapmeta{Half-life: durable. Archetypes: transport, scaling, failure. Exercise target: 30. Project track: physical build plus instrumented measurement (hybrid). Hazard class: medium (laser pointers Class~2, Class~3R; safety-glass and beam-block protocols required). Reader path default: standard, with sections explicitly tagged. Named cases: Hubble Space Telescope primary mirror (1990).}

\noindent
Optics is electromagnetism at wavelengths the human eye can see and at
nearby wavelengths the engineer can manipulate with the same
instruments. The discipline splits into geometric optics (rays
travelling in straight lines, refracting through lenses, reflecting
from mirrors) and wave optics (interference, diffraction, coherence,
polarisation). Both descriptions are correct at the wavelengths and
geometries the engineer cares about; geometric optics is faster when
the relevant geometry is much larger than $\lambda$, wave optics
takes over when geometry approaches $\lambda$. The chapter develops
both, ending with the canonical optical failure mode of the late
twentieth century.

\section{Geometric optics: lenses and mirrors}\pathtag{core}

A \engterm{ray} is the trajectory along which optical energy
propagates in the limit that the wavelength is much smaller than every
relevant geometric scale. Rays bend at refractive interfaces by
Snell's law (Chapter 10), $n_{1}\sin\theta_{1} = n_{2}\sin\theta_{2}$,
reflect at mirrors by the equal-angle rule,
and propagate in straight lines through homogeneous media. Figure
\ref{fig:vol04ch11:snell-refraction} shows the single-interface case
that every lens and prism is built from. Geometric
optics is the working description for camera lenses, telescope
mirrors, microscope objectives, and projection optics; it fails when
diffraction or interference becomes visible, typically when an
aperture's diameter approaches a few wavelengths.

\input{volumes/04-energy/ch11-optics/figures/fig-snell-refraction.tex}

A \engterm{thin lens} is an optical element whose thickness is small
compared to its focal length. The thin-lens equation relates object
distance $s_{o}$ (from the lens), image distance $s_{i}$, and focal
length $f$:
\[
\frac{1}{s_{o}} + \frac{1}{s_{i}} = \frac{1}{f}.
\]
Sign conventions: $s_{o}$ positive for real objects on the incoming
side, $s_{i}$ positive for real images on the outgoing side,
$f$ positive for converging lenses, negative for diverging. The
\engterm{lateral magnification} is
\[
M = -\frac{s_{i}}{s_{o}},
\]
with the negative sign capturing the inversion of a real image cast
by a converging lens. Figure \ref{fig:vol04ch11:lens-ray-diagram}
traces the three principal rays that locate the image of a point and
fixes the sign conventions in a single picture. The calculator
\texttt{code/thin\_lens.py} returns image distance, magnification, and
orientation for any object distance and focal length.

\input{volumes/04-energy/ch11-optics/figures/fig-lens-ray-diagram.tex}

The lens-maker's equation gives the focal length of a thin lens of
refractive index $n$ with surface radii $R_{1}$ (first surface) and
$R_{2}$ (second surface),
\[
\frac{1}{f} = (n-1)\left(\frac{1}{R_{1}} - \frac{1}{R_{2}}\right),
\]
under the conventions that $R$ is positive when the centre of
curvature is on the outgoing side. A biconvex lens of radii $R_{1} =
+50\,\si{\milli\meter}$, $R_{2} = -50\,\si{\milli\meter}$ in BK7 glass
($n = 1.52$) has $f = 1/[0.52\times(1/50 + 1/50)] = 48\,\si{\milli
\meter}$. The result is the working tool for first-order optical
design; the actual lens will deviate from the thin-lens prediction
by terms in $(s_{o}/R)^{2}$ and higher (Section 11.2 on aberrations).
The refractive indices and Abbe numbers of the common optical glasses,
the raw material of every lens design, are tabulated in
\texttt{data/glass\_indices.csv}.

A \engterm{spherical mirror} obeys the same equation as a thin lens
in absolute value but with a sign convention specific to reflective
geometry: $f = R/2$ for a concave mirror, with $R$ the radius of
curvature. A concave mirror of radius $1\,\si{\meter}$ has focal
length $0.5\,\si{\meter}$ and produces a real, inverted image of a
distant object at its focal plane. A convex mirror has negative
focal length and produces a virtual, erect, demagnified image used in
car side mirrors and in many security applications. Figure
\ref{fig:vol04ch11:mirror-imaging} traces the image of an object beyond
the centre of curvature of a concave mirror; the construction parallels
the lens diagram, with reflection replacing refraction.

\input{volumes/04-energy/ch11-optics/figures/fig-mirror-imaging.tex}

The \engterm{numerical aperture} of an optical system is
\[
\text{NA} = n\sin\theta_{\max},
\]
where $\theta_{\max}$ is the half-angle of the cone of light the
system accepts and $n$ is the refractive index of the working medium
(air, typically; oil immersion in high-NA microscopy). The
$f$-number is the inverse working approximation: $f/\# = 1/(2\,\text{
NA})$ at small angles. A camera lens at $f/2.8$ has NA $\approx 0.18$;
a microscope objective at NA $= 1.4$ (oil immersion) has an
$f/\#$ of about $0.36$.

\section{Ray-transfer (ABCD) matrices}\pathtag{core}

The thin-lens equation handles one element at a time, and chaining it through a
multi-element system by hand grows error-prone the moment a third surface
enters. Paraxial optics has a bookkeeping that scales: represent each ray at a
reference plane by the two numbers that fix it under the small-angle
approximation, its height $y$ above the optical axis and its angle $\theta$ to
the axis, and write any optical operation as a $2\times 2$ matrix acting on the
column vector $(y, \theta)^{\mathsf{T}}$. The matrix is the \engterm{ray-transfer
matrix} or ABCD matrix,
\[
\begin{pmatrix} y' \\ \theta' \end{pmatrix}
= \begin{pmatrix} A & B \\ C & D \end{pmatrix}
  \begin{pmatrix} y \\ \theta \end{pmatrix},
\]
and the entire discipline of first-order optical layout reduces to multiplying
these matrices in the order the ray meets the elements.

Two operations generate every paraxial system. Free propagation over a distance
$d$ in a medium of uniform index leaves the angle unchanged and raises the
height by $d\theta$, so
\[
\mathbf{M}_{\text{prop}}(d) = \begin{pmatrix} 1 & d \\ 0 & 1 \end{pmatrix}.
\]
Refraction at a thin lens of focal length $f$ leaves the height unchanged at the
lens plane and bends the angle by $-y/f$, the defining action of a lens on a
paraxial ray, so
\[
\mathbf{M}_{\text{lens}}(f) = \begin{pmatrix} 1 & 0 \\ -1/f & 1 \end{pmatrix}.
\]
A flat dielectric interface from index $n_1$ to $n_2$ preserves the height and
scales the angle by $n_1/n_2$ (Snell's law in the small-angle limit), and a
spherical refracting surface of radius $R$ carries the matrix
$\big(\begin{smallmatrix} 1 & 0 \\ (n_1-n_2)/(n_2 R) & n_1/n_2
\end{smallmatrix}\big)$, from which the thin-lens matrix follows by composing
the two surfaces of a lens in air. A curved mirror of radius $R$ acts as a lens
of focal length $R/2$ in the unfolded diagram. The matrix of a system is the
ordered product of its element matrices, written right to left so that the first
element the ray meets sits rightmost,
\[
\mathbf{M}_{\text{sys}} = \mathbf{M}_n \cdots \mathbf{M}_2\,\mathbf{M}_1 .
\]
For a ray in air entering and leaving the system, every generating matrix has
unit determinant, so the system matrix does too; $\det\mathbf{M} = n_{\text{in}}
/n_{\text{out}}$ in general, a conserved quantity that catches arithmetic errors
the moment a product fails it.

The four elements read directly as system properties. The system focal length is
$f_{\text{sys}} = -1/C$, so the off-diagonal lower-left element is the optical
power up to a sign. The imaging condition, that an object plane maps to an image
plane, is the requirement that the $B$ element of the full object-to-image matrix
vanish, $B = 0$; when it does, the $A$ element is the lateral magnification
$M = A$ and the $D$ element is the angular magnification. The reason is direct:
with $\mathbf{M}_{\text{obj}\to\text{img}} = \big(\begin{smallmatrix} A & B \\ C
& D\end{smallmatrix}\big)$ the image height is $y' = A y + B\theta$, and a sharp
image requires that all rays leaving one object point ($y$ fixed, $\theta$
ranging) arrive at one image height independent of $\theta$, which forces
$B = 0$ and leaves $y' = A y$, magnification $A$.

\paragraph{Worked example: the thin-lens equation from the matrix.} Compose
propagation $s_o$, the lens, and propagation $s_i$ to confirm that the matrix
method reproduces the elementary result and to see the imaging condition do its
work. The object-to-image matrix is
\[
\mathbf{M} = \begin{pmatrix} 1 & s_i \\ 0 & 1\end{pmatrix}
\begin{pmatrix} 1 & 0 \\ -1/f & 1\end{pmatrix}
\begin{pmatrix} 1 & s_o \\ 0 & 1\end{pmatrix}
= \begin{pmatrix} 1 - s_i/f & s_o + s_i - s_o s_i/f \\
-1/f & 1 - s_o/f \end{pmatrix}.
\]
The imaging condition $B = 0$ reads $s_o + s_i - s_o s_i/f = 0$; dividing by
$s_o s_i$ gives $1/s_i + 1/s_o - 1/f = 0$, the thin-lens equation. At that
focus the upper-left element is $A = 1 - s_i/f$, which the imaging condition
turns into $A = 1 - s_i(1/s_o + 1/s_i) = -s_i/s_o = M$, the magnification with
its inversion sign. The lower-left element is $C = -1/f$ throughout, confirming
$f_{\text{sys}} = -1/C = f$ for a single lens. The matrix recovers every
elementary result and, unlike the scalar equation, carries straight over to a
system of arbitrary element count.

\paragraph{Worked example: a two-lens relay and its principal planes.} Take a
$+60\,\si{\milli\meter}$ lens and a $+40\,\si{\milli\meter}$ lens separated by
$d = 30\,\si{\milli\meter}$. The system matrix from the first lens plane to the
second is
\[
\mathbf{M} = \begin{pmatrix} 1 & 0 \\ -1/40 & 1\end{pmatrix}
\begin{pmatrix} 1 & 30 \\ 0 & 1\end{pmatrix}
\begin{pmatrix} 1 & 0 \\ -1/60 & 1\end{pmatrix}.
\]
Carrying the product, the propagation-lens product is
$\big(\begin{smallmatrix} 1 & 30 \\ -1/60 & 1/2\end{smallmatrix}\big)$, and
multiplying by the second lens gives
$C = -1/40\cdot 1 - 1/60 \cdot 1 + (-1/40)(30)(-1/60)$, which evaluates to
\[
C = -\frac{1}{40} - \frac{1}{60} + \frac{30}{2400}
= -0.025 - 0.01667 + 0.0125 = -0.02917\,\si{\per\milli\meter},
\]
so $f_{\text{sys}} = -1/C = 34.3\,\si{\milli\meter}$, shorter than either lens
because the two converging elements combine. The Gullstrand relation
$1/f_{\text{sys}} = 1/f_1 + 1/f_2 - d/(f_1 f_2) = 1/60 + 1/40 - 30/2400 =
0.02917$ gives the same number, the matrix and the scalar formula in agreement.
The principal planes, the reference planes from which $f_{\text{sys}}$ is
measured, sit a distance $x_H = (A-1)/C$ in front of the first lens and
$x_{H'} = (1-D)/C$ behind the second; the matrix yields them at no extra cost,
where the scalar formula needs separate construction.

\paragraph{Worked example: laser-cavity stability from the round-trip matrix.}
A two-mirror laser resonator of length $L$ with mirror radii $R_1$ and $R_2$
supports a stable transverse mode only when a ray launched near the axis stays
near the axis after arbitrarily many round trips. The round-trip matrix is the
ordered product of propagation $L$, reflection from mirror~2 (focal length
$R_2/2$), propagation $L$, and reflection from mirror~1. Define the
$g$-parameters $g_i = 1 - L/R_i$; the round-trip matrix has trace $A + D =
4 g_1 g_2 - 2$. A ray iterated by a unimodular matrix stays bounded only when
the eigenvalues lie on the unit circle, which requires $|A + D|/2 \le 1$, hence
\[
0 \le g_1 g_2 \le 1 .
\]
This is the resonator stability condition. Figure
\ref{fig:vol04ch11:abcd-cavity-stability} maps it in the $(g_1, g_2)$ plane and
locates the named designs. A symmetric confocal cavity ($R_1 = R_2 = L$) sits at
the origin, deep inside the stable region with wide alignment tolerance; a
plane-parallel cavity ($R \to \infty$, $g = 1$) sits on the corner of the
boundary, where any misalignment walks the mode off the mirrors, which is why
plane-parallel resonators are notoriously hard to align. The stability map is a
ray-transfer-matrix result that no scalar imaging formula reaches.

\input{volumes/04-energy/ch11-optics/figures/fig-abcd-cavity-stability.tex}

\begin{archetype}[Composition and conservation]
Ray-transfer matrices turn optical-system design into matrix multiplication, the
same move that turns a chain of mechanical or electrical two-ports into a product
of transfer matrices. Two properties recur across the book's two-port methods.
Composition: the system matrix is the ordered product of element matrices.
Conservation: the determinant is fixed (here $n_{\text{in}}/n_{\text{out}}$,
unity for a system in air), and a product that violates it is wrong. The same
structure appears in the thin-film transfer matrices of the next section, in the
ABCD-law propagation of Gaussian beams, and in the two-port networks of Volume
VII.
\end{archetype}

\section{Aberrations}\pathtag{standard}

Real optical systems deviate from the paraxial (thin-lens, small-
angle) ideal because the spherical surfaces, polychromatic light,
and finite apertures of practical optics produce systematic
distortions called \engterm{aberrations}. Five monochromatic
aberrations are catalogued by the Seidel classification, and two
chromatic aberrations complete the list.

\engterm{Spherical aberration} is the failure of a spherical surface
to focus all parallel rays to a single point: rays striking the
surface farther from the axis focus closer to the lens than paraxial
rays. The transverse spherical aberration scales as $r^{3}$ in the
ray's radial position at the lens. The cure is either an aspheric
surface (a paraboloid focuses all parallel rays to a single point
exactly, used for telescope primary mirrors and high-quality camera
lenses) or a combination of glass elements whose spherical
aberrations cancel by sign.

\engterm{Coma} appears for off-axis points: the image of a point
becomes a comet-shaped streak with its tail pointing radially outward.
Coma scales linearly with field angle and quadratically with aperture.
It is corrected by combining lenses of different shapes (the Abbe
sine condition) or by using off-axis paraboloidal mirrors arranged so
that the on-axis ray reaches the focal point at the correct phase.

\engterm{Astigmatism} of an optical system (distinct from the
eye-anatomical astigmatism) is the failure of a system to focus
tangential and sagittal rays to the same point at off-axis field
positions; tangential rays focus closer to the lens than sagittal
rays, producing an elongated image. Cylindrical lenses introduce or
correct astigmatism deliberately.

\engterm{Field curvature} (Petzval curvature) is the failure of a
flat object plane to image to a flat image plane; the image surface
is a curved bowl. The Petzval condition for flatness requires that
the sum of $1/(n_{i}f_{i})$ over all lens elements be zero, achievable
with a careful selection of converging and diverging elements.

\engterm{Distortion} is the failure of magnification to be constant
across the field; barrel distortion squashes the corners inward,
pincushion distortion stretches them outward. Modern digital cameras
often correct distortion in software, but the optical aberration is
present in the raw image.

\engterm{Chromatic aberration} is the wavelength dependence of the
refractive index: shorter wavelengths (blue) focus closer to the
lens, longer (red) farther. Axial chromatic aberration is the
focal-length variation with wavelength; transverse chromatic
aberration is the magnification variation with wavelength. The cure
is an \engterm{achromatic doublet}: a converging crown-glass element
and a diverging flint-glass element bonded together, with the
combination's chromatic aberration cancelled at two wavelengths
(typically Fraunhofer F at $486\,\si{\nano\meter}$ and C at $656\,\si{
\nano\meter}$). An apochromat extends correction to three wavelengths;
the price is more glass elements and tighter tolerances.

\begin{archetype}[Scaling and failure]
Every aberration scales as a power of $r$ (the ray's radial position
on the lens) and a power of $\theta$ (the field angle). Spherical
aberration is $r^{3}$ at all field positions; coma is $r^{2}\theta$;
astigmatism and field curvature are $r\theta^{2}$; distortion is
$\theta^{3}$. The engineer who knows the scaling can predict which
aberrations dominate at given aperture and field-of-view and direct
the design effort to the dominant terms.
\end{archetype}

\begin{mastery}
\textbf{The Strehl ratio and the Mar\'echal criterion.} A single number
grades how close a real optical system comes to its diffraction limit:
the \engterm{Strehl ratio} $S$, the on-axis peak intensity of the
actual point-spread function divided by the peak of the aberration-free
Airy pattern. For small wavefront errors the extended Mar\'echal
approximation gives $S \approx \exp(-(2\pi\sigma/\lambda)^{2})$, where
$\sigma$ is the root-mean-square wavefront error. The conventional
threshold for ``diffraction-limited'' performance is $S \ge 0.80$, which
the formula reaches at $\sigma = \lambda/14$, the Mar\'echal criterion.
A system at the classical quarter-wave peak-to-valley tolerance sits at
roughly $S = 0.80$ for smooth low-order error. The Hubble primary mirror
of the failure section had a wavefront error near $\lambda/2$ RMS, which
drives $S$ below $0.2$: the formula turns the polishing error directly
into the lost contrast that the first images showed.
\end{mastery}

\section{Wave optics: interference and diffraction}\pathtag{core}

When the wavelength approaches the scale of any obstacle or aperture
in the system, the geometric-optics picture must be extended to
include wave effects. Two phenomena anchor the wave-optics chapter:
interference and diffraction.

\engterm{Interference} occurs when two coherent waves superpose. The
combined intensity is
\[
I = I_{1} + I_{2} + 2\sqrt{I_{1}I_{2}}\cos\delta,
\]
where $\delta$ is the phase difference between the two waves. For
equal-amplitude waves, $I = 4I_{0}\cos^{2}(\delta/2)$, with maxima at
$\delta = 0, 2\pi, 4\pi, \ldots$ and zero at $\delta = \pi, 3\pi,
\ldots$ The \engterm{double-slit experiment} of Thomas Young (1801)
produced the canonical demonstration; the fringe spacing on a screen
at distance $L$ from two slits of separation $d$ is
\[
\Delta y = \frac{\lambda L}{d},
\]
for $\lambda \ll d$. The same fringe-spacing formula governs the
working of every interferometer in metrology. Figure
\ref{fig:vol04ch11:young-double-slit} shows the geometry and the
resulting $\cos^{2}$ intensity pattern.

\input{volumes/04-energy/ch11-optics/figures/fig-young-double-slit.tex}

A \engterm{thin film} produces interference of a different geometry:
the two interfering beams are the reflection from the film's top surface
and the reflection from its bottom surface, separated by the optical
path $2 n_{f} t\cos\theta_{f}$ through a film of index $n_{f}$ and
thickness $t$ (Figure \ref{fig:vol04ch11:thin-film}). A reflection at a
boundary from a lower index to a higher index adds a $\lambda/2$ phase
shift, which the engineer must track when counting orders. The same
arithmetic that produces the colours of a soap film designs an
anti-reflection coating: a quarter-wave layer of index
$\sqrt{n_{\text{out}} n_{\text{sub}}}$ makes the two reflections cancel
at the design wavelength. The routine \texttt{code/thin\_film.py}
computes the path difference and the quarter-wave thickness.

\input{volumes/04-energy/ch11-optics/figures/fig-thin-film.tex}

The two-beam picture of a single film is exact only when the two surface
reflections are weak; a real coating has many internal bounces, and a stack of
layers has more interfaces than two beams can track. The bookkeeping that scales
to any number of layers is the \engterm{characteristic matrix} method, the
thin-film cousin of the ray-transfer matrix. Represent the field in each medium
by the tangential electric and magnetic field amplitudes $(E, H)$ at a plane;
crossing a homogeneous layer of index $n$, physical thickness $d$, and
phase thickness $\delta = (2\pi/\lambda)\,n d\cos\theta$ multiplies the field
vector by the layer's characteristic matrix
\[
\mathbf{M}_{\text{layer}}
= \begin{pmatrix} \cos\delta & (i/\eta)\sin\delta \\
i\,\eta\sin\delta & \cos\delta \end{pmatrix},
\]
where $\eta = n\cos\theta$ is the layer's optical admittance at normal incidence
(the simple form $\eta = n$). The matrix of a multilayer stack is the ordered
product of its layers' matrices, exactly as for rays,
$\mathbf{M} = \mathbf{M}_1 \mathbf{M}_2 \cdots \mathbf{M}_N$. Each layer matrix
has determinant $\cos^2\delta + \sin^2\delta = 1$, so the stack matrix is
unimodular, the conservation check again. The reflectance follows from the
product matrix and the substrate admittance $\eta_s$ and incident-medium
admittance $\eta_0$: writing
$\big(\begin{smallmatrix} B \\ C\end{smallmatrix}\big) = \mathbf{M}
\big(\begin{smallmatrix} 1 \\ \eta_s\end{smallmatrix}\big)$, the amplitude
reflection coefficient is
\[
r = \frac{\eta_0 B - C}{\eta_0 B + C},
\qquad R = |r|^2 .
\]
The whole multilayer-coating industry, anti-reflection coats, dielectric mirrors,
bandpass filters, and the wavelength-locking filters of the fibre case study,
runs on this one recursion.

\paragraph{Worked example: the quarter-wave layer recovers the geometric-mean
rule.} A single quarter-wave layer has $\delta = \pi/2$, so $\cos\delta = 0$ and
$\sin\delta = 1$, collapsing its matrix to
$\big(\begin{smallmatrix} 0 & i/\eta_f \\ i\,\eta_f & 0\end{smallmatrix}\big)$.
Then $B = i/\eta_f \cdot \eta_s$ scaled, and carrying the algebra the reflectance
vanishes when $\eta_f^2 = \eta_0\,\eta_s$, the admittance condition that for
normal incidence reads $n_f = \sqrt{n_0 n_s}$, the geometric-mean rule the
two-beam argument quoted. The matrix method earns its place not on this case,
which the simple picture already handled, but on what the simple picture cannot
reach: a two-layer V-coat that nulls the reflectance at a chosen wavelength with
realisable indices, and the high-reflector stack below.

\paragraph{Worked example: a quarter-wave high-reflector stack.} Stack $N$
pairs of a high-index quarter-wave layer ($n_H$) and a low-index quarter-wave
layer ($n_L$) on a substrate. Each quarter-wave layer at the design wavelength
contributes the off-diagonal matrix above, and a high-low pair multiplies to the
diagonal matrix $\big(\begin{smallmatrix} -n_L/n_H & 0 \\ 0 &
-n_H/n_L\end{smallmatrix}\big)$. Raising it to the $N$th power and feeding the
result through the reflectance formula gives a reflectance that climbs toward
unity as
\[
R = \left[\frac{1 - (n_H/n_L)^{2N}(n_H^2/n_s)}
{1 + (n_H/n_L)^{2N}(n_H^2/n_s)}\right]^{2},
\]
so each added pair multiplies the residual transmission by $(n_L/n_H)^2$. With
titanium dioxide ($n_H = 2.4$) and silica ($n_L = 1.46$) the per-pair factor is
$(1.46/2.4)^2 = 0.37$, so ten pairs reach $R$ above $99.9\,\%$, the dielectric
mirror that outperforms any metal coating and carries the megawatt
circulating power of a laser cavity without absorbing enough to melt. Figure
\ref{fig:vol04ch11:multilayer-reflectance} contrasts the single-layer
anti-reflection dip, the bare-glass baseline, and this high-reflector stop band,
all three from the same matrix recursion run with different layer choices. The
stop-band width narrows and the reflectance rises as the index contrast and the
pair count grow, the two design handles a coating engineer turns.

\input{volumes/04-energy/ch11-optics/figures/fig-multilayer-reflectance.tex}

\engterm{Diffraction} is the spreading of a wave after passing
through an aperture or around an edge. A wave of wavelength $\lambda$
passing through a slit of width $a$ produces a single-slit
diffraction pattern with first dark fringe at angle
\[
\sin\theta = \frac{\lambda}{a},
\]
and a half-width on a screen at distance $L$ of $\lambda L/a$. The
full intensity profile is the $[\sin\beta/\beta]^{2}$ pattern of Figure
\ref{fig:vol04ch11:single-slit-diffraction}, with a central maximum
twice as wide as the side lobes. A periodic array of slits, a
\engterm{diffraction grating} of period $d$, concentrates the
transmitted light into discrete orders at angles
$d\sin\theta_{m} = m\lambda$ (Figure \ref{fig:vol04ch11:grating-orders});
because the order angle grows with wavelength, a grating disperses white
light into a spectrum and is the working element of most spectrometers.

\input{volumes/04-energy/ch11-optics/figures/fig-single-slit-diffraction.tex}

The $[\sin\beta/\beta]^{2}$ form is worth deriving rather than quoting,
because the derivation fixes the meaning of every factor and shows why
the central lobe is twice the width of the side lobes. Divide the slit of
width $a$ into strips of width $dx$ at position $x$ measured from the
slit centre. Each strip radiates a spherical wavelet; in the Fraunhofer
(far-field) limit the strips are illuminated in phase, and the wavelet
reaching a screen point at angle $\theta$ from strip $x$ carries an extra
path $x\sin\theta$ relative to the wavelet from the centre, hence an
extra phase $k x\sin\theta$ with $k = 2\pi/\lambda$. Summing the
phasors across the slit gives the field amplitude
\[
E(\theta) \propto \int_{-a/2}^{+a/2} e^{\,i k x\sin\theta}\,dx
= a\,\frac{\sin\!\big(\tfrac{1}{2}k a\sin\theta\big)}
{\tfrac{1}{2}k a\sin\theta}.
\]
Writing the half-phase across the slit as
$\beta \equiv \tfrac{1}{2}k a\sin\theta = (\pi a/\lambda)\sin\theta$, the
amplitude is $E \propto a\,(\sin\beta)/\beta$ and the intensity is its
square,
\[
I(\theta) = I_{0}\left[\frac{\sin\beta}{\beta}\right]^{2}.
\]
The central maximum sits at $\beta = 0$, where $\sin\beta/\beta \to 1$ by
the small-angle limit, so $I = I_{0}$. The pattern vanishes wherever
$\sin\beta = 0$ with $\beta \ne 0$, that is at $\beta = m\pi$ for
non-zero integer $m$, which is $\sin\theta = m\lambda/a$; the first
minimum at $m = 1$ recovers $\sin\theta = \lambda/a$. The central lobe
spans $-\pi < \beta < +\pi$, a phase range of $2\pi$, while each side lobe
spans a phase range of $\pi$ between successive zeros, which is the origin
of the central lobe's doubled width. The side-lobe peaks fall near
$\beta = (m + \tfrac{1}{2})\pi$, and the first of them carries
$[\sin(3\pi/2)/(3\pi/2)]^{2} = (2/3\pi)^{2} = 0.045$, about
$4.7\,\%$ of the central peak, the stray-light floor that limits the
contrast of any single-aperture imaging system. Figure
\ref{fig:vol04ch11:single-slit-phasor} plots the intensity against
$\beta$ and marks these features.

\input{volumes/04-energy/ch11-optics/figures/fig-single-slit-phasor.tex}

The Fraunhofer result above assumes the screen is far enough that the
strips radiate in phase and the path difference is linear in $x$. The
boundary between this far-field regime and the near-field
(\engterm{Fresnel}) regime, where the quadratic term in the path
expansion still matters, is set by the \engterm{Fresnel number}
$N_{F} = a^{2}/(\lambda L)$ for an aperture of size $a$ at distance $L$.
Fraunhofer diffraction holds when $N_{F} \ll 1$, Fresnel diffraction
when $N_{F} \gtrsim 1$. A $1\,\si{\milli\meter}$ slit at
$\lambda = 550\,\si{\nano\meter}$ reaches $N_{F} = 1$ at
$L = a^{2}/\lambda = (10^{-3})^{2}/(5.5\times10^{-7}) = 1.8\,\si{\meter}$,
so a bench demonstration at a few metres is comfortably in the
Fraunhofer regime, while the same slit viewed at $10\,\si{\centi\meter}$
shows the structured Fresnel pattern instead of the clean sinc. The
lens trick that every spectrometer uses is to place the observation
plane at the focal plane of a lens, which images the angular
Fraunhofer pattern at any working distance because the lens performs the
far-field Fourier transform optically.

\input{volumes/04-energy/ch11-optics/figures/fig-grating-orders.tex}

A
circular aperture of diameter $D$ produces an Airy disc with first
dark ring at
\[
\sin\theta = 1.22\frac{\lambda}{D}.
\]
The $1.22$ factor is the first zero of the Bessel function $J_{1}$,
the canonical content of two-dimensional aperture diffraction. The
Airy disc is the angular resolution limit of every telescope, camera,
and microscope; a $200\,\si{\milli\meter}$ telescope at $550\,\si{
\nano\meter}$ resolves angular separations of $3.4\,\si{\micro
\radian}$ ($0.7$ arc-second), the diffraction-limited resolution of
the instrument.

The \engterm{Rayleigh criterion} states that two point sources are
just resolved when the Airy disc of one falls on the first dark ring
of the other; this gives the angular resolution
\[
\theta_{\min} = 1.22\frac{\lambda}{D}.
\]
The criterion is conservative; more sophisticated criteria
(Sparrow, Dawes) recognise smaller resolved separations. The
engineer's working number is the Rayleigh value. Figure
\ref{fig:vol04ch11:rayleigh-resolution} shows two point sources at the
limit, where the summed intensity dips about a quarter below the peaks.
The aperture-resolution trade is tabulated across instruments from the
human eye to the Extremely Large Telescope in
\texttt{data/telescope\_aperture\_resolution.csv}, and the diffraction
calculator \texttt{code/diffraction.py} returns the Rayleigh angle and
the grating orders for any aperture and wavelength.

\input{volumes/04-energy/ch11-optics/figures/fig-rayleigh-resolution.tex}

\begin{estimation}
\textbf{Question.} A spy satellite orbits at $300\,\si{\kilo\meter}$.
What primary-mirror diameter does it need to read a car licence plate
from orbit, and is that diameter physically plausible?

\textbf{Working.} Reading a licence plate means resolving the gaps
between characters, features about $1\,\si{\centi\meter}$ across at a
range of $R = 300\,\si{\kilo\meter}$. The required angular resolution is
$\theta = (0.01\,\si{\meter})/(3\times10^{5}\,\si{\meter}) = 3.3\times
10^{-8}\,\si{\radian}$. The Rayleigh criterion
$\theta = 1.22\lambda/D$ at a visible wavelength
$\lambda = 550\,\si{\nano\meter}$ inverts to
\[
D = \frac{1.22\lambda}{\theta} = \frac{1.22\times 5.5\times10^{-7}}
{3.3\times10^{-8}} \approx 20\,\si{\meter}.
\]
A $20\,\si{\meter}$ space mirror exceeds anything yet flown (Hubble is
$2.4\,\si{\meter}$, the James Webb Space Telescope $6.5\,\si{\meter}$
and segmented), so single-character licence-plate reading from low
orbit is past the diffraction limit of any plausible single aperture.
Relaxing the target to vehicle identification (features
$\sim 0.5\,\si{\meter}$) drops the required diameter to about
$0.4\,\si{\meter}$, which is comfortably within reach, and explains why
real reconnaissance satellites resolve vehicles and structures but not
text. Atmospheric turbulence, neglected here, only makes the text case
worse. The order-of-magnitude verdict, that diffraction alone forbids
the cinematic licence-plate read, holds under every reasonable
assumption.
\end{estimation}

\section{Coherence and lasers (preview)}\pathtag{standard}

Two waves are \engterm{coherent} if they maintain a fixed phase
relationship over time. Coherence is measured along two axes:
temporal coherence (length over which phase is predictable) and
spatial coherence (transverse extent over which phase is predictable).
A thermal source has short coherence in both senses; a laser has
long coherence in both senses, which is why lasers produce
interference fringes with macroscopic path differences and why
thermal sources do not without a beam-splitter and short paths.

The \engterm{coherence length} $L_{c} = c\tau_{c}$ is the distance
over which the wave remains coherent in time. For a source of
linewidth $\Delta f$,
\[
L_{c} \approx \frac{c}{\Delta f}.
\]
A laboratory helium-neon laser has $\Delta f \approx 1.5\,\si{\giga
\hertz}$ (without frequency stabilisation) and $L_{c} \approx 20\,\si{
\centi\meter}$; a frequency-stabilised iodine-locked HeNe has $\Delta
f \le 10\,\si{\kilo\hertz}$ and $L_{c} \ge 30\,\si{\kilo\meter}$. A
fluorescent lamp emits at multiple atomic transitions with $\Delta f$
of order $10^{12}\,\si{\hertz}$ and $L_{c} \approx 300\,\si{\micro
\meter}$; a thermal incandescent bulb has $L_{c}$ of order microns.

The coherence length is the working specification that decides which source an
interferometer can use. A Michelson interferometer forms fringes only while
its two arms differ by less than the coherence length, so a metrology
instrument measuring a metre-scale displacement needs a source with a
coherence length of metres, which an unstabilised gas laser barely provides
and a frequency-stabilised one provides with room to spare. The same number
read the other way is a feature: optical coherence tomography deliberately
uses a broadband, short-coherence source so that fringes form only from a
thin slice of tissue at the matched path depth, and scanning the reference arm
sweeps that slice through the sample to build a depth profile. Short coherence
is a defect for a length-measuring interferometer and the operating principle
of a depth-resolving one.

The \engterm{laser} (Light Amplification by Stimulated Emission of
Radiation) produces coherent light through stimulated emission in a
gain medium, with feedback provided by a pair of mirrors forming an
optical cavity. The three operating regimes are continuous-wave (the
laser produces a constant output beam), $Q$-switched (the cavity is
held lossy until a population inversion builds, then opened, producing
a giant pulse of $\sim 10\,\si{\nano\second}$ duration), and
mode-locked (multiple cavity modes are forced to oscillate in phase,
producing femtosecond pulses). Working applications: continuous-wave
diode lasers for fibre communications, $Q$-switched solid-state
lasers for laser-induced breakdown spectroscopy, mode-locked
Ti:sapphire lasers for two-photon microscopy and laser materials
processing. The detailed physics is in Chapter 12, where laser
operation is treated as a quantum-mechanical device problem.

\section{Imaging systems: cameras, microscopes, telescopes}\pathtag{standard}

A \engterm{camera} is an imaging system that forms an image of an
extended object on a detector (film or CCD). The camera's specification
is the focal length, the aperture, and the detector size. Focal length
sets the angle of view: a $50\,\si{\milli\meter}$ lens on a
$36\times 24\,\si{\milli\meter}$ full-frame sensor has horizontal
angle of view $\approx 40^{\circ}$, the standard ``normal'' lens; a
$24\,\si{\milli\meter}$ lens gives $74^{\circ}$ (wide-angle); a
$200\,\si{\milli\meter}$ lens gives $10^{\circ}$ (telephoto). The
aperture (relative aperture $f/\#$) sets the working depth of field
and the diffraction limit; an $f/2$ aperture diffraction-limits at
about $1.4\,\si{\micro\meter}$ Airy-disc radius at $550\,\si{\nano
\meter}$. Modern smartphone cameras have aperture $f/1.6$ and pixel
pitch $\approx 1\,\si{\micro\meter}$; the camera is diffraction-
limited rather than sensor-limited at full aperture, and the trick
of computational photography is to recover the lost resolution by
combining multiple exposures with sub-pixel offsets.

A \engterm{microscope} is an imaging system optimised for high
magnification of small objects. The two stages are the objective
(forms an intermediate image at $\sim 160\,\si{\milli\meter}$ from
the back focal plane) and the eyepiece (forms a magnified virtual
image of the intermediate image). The system's resolution is set by
the objective's NA:
\[
d_{\min} = 0.61\frac{\lambda}{\text{NA}}.
\]
At NA $= 1.4$ (oil immersion) and $\lambda = 500\,\si{\nano\meter}$,
$d_{\min} = 220\,\si{\nano\meter}$. Sub-wavelength microscopy
techniques (STED, PALM/STORM, structured illumination) circumvent the
diffraction limit by exploiting nonlinear or molecular-scale
phenomena; their resolution is no longer set by the diffraction limit
of the imaging system but by the spatial precision of the
illumination or the labelling.

A \engterm{telescope} is an imaging system optimised for high angular
resolution of distant objects. The objective (primary mirror or
objective lens) forms an image at its focal plane; the eyepiece
magnifies. The angular resolution is set by the primary's diameter
$D$ via the Rayleigh criterion. The collecting area scales as $D^{2}$;
the resolving power scales as $D$; both are reasons large telescopes
are large.

The largest ground-based optical telescopes have primary mirrors of
$8$ to $10\,\si{\meter}$ diameter (the VLT, Subaru, Keck I and II);
the next generation (Extremely Large Telescope, expected first light
in the late 2020s) will have a $39\,\si{\meter}$ segmented primary.
Atmospheric seeing limits ground-based telescopes to $\sim 1$
arc-second resolution without adaptive optics; with adaptive optics
(a deformable mirror corrects the atmospheric phase distortion in
real time at $\sim 1\,\si{\kilo\hertz}$), large telescopes can reach
their diffraction-limited resolution at the best sites. Space-based
telescopes (Hubble, JWST) avoid atmospheric distortion entirely; the
trade-off is their cost and the difficulty of repair (in Hubble's
case, four servicing missions; in JWST's case, none possible).

\section{Fourier optics and the transfer function}\pathtag{standard}

Geometric optics treats an image as a map of points; wave optics, carried to its
conclusion, treats an image as a filtered version of the object's spatial
content. The bridge is the result that a lens performs a Fourier transform. A
field $E(x,y)$ in the front focal plane of a lens of focal length $f$ produces in
the back focal plane the field
\[
E_f(u,v) \propto \widetilde{E}\!\left(\frac{u}{\lambda f},
\frac{v}{\lambda f}\right),
\]
where $\widetilde{E}$ is the two-dimensional Fourier transform of the input and
the back-focal-plane coordinates $(u,v)$ scale to spatial frequencies
$f_x = u/(\lambda f)$, $f_y = v/(\lambda f)$. The far-field Fraunhofer pattern
the diffraction section derived is the same transform; the lens brings the
infinitely distant far field to a reachable plane, which is why every
spectrometer and every coherent imaging system places a detector at a lens focal
plane. A point at radius $\rho$ in the back focal plane corresponds to a spatial
frequency $\rho/(\lambda f)$ in the object, so the back focal plane is a physical
map of the object's spatial-frequency content, the \engterm{Fourier plane}.

A two-lens \engterm{4f system}, two lenses of focal length $f$ spaced $2f$ apart
with the object one focal length before the first and the image one focal length
after the second, transforms to the Fourier plane at the shared focus and
transforms back to an image. Placing a mask in the Fourier plane multiplies the
object's spatial spectrum, so the system filters the image in hardware: a small
central stop removes the low frequencies and sharpens edges (a high-pass filter),
a central pinhole passes only the low frequencies and smooths the image (a
low-pass filter, the spatial-filtering pinhole that cleans a laser beam), and a
knife edge at the Fourier plane converts phase variations into intensity, the
Schlieren and phase-contrast methods that make transparent flows and unstained
cells visible. The Fourier plane is where an optical system's information lives in
a form the engineer can edit.

The same picture grades how faithfully a system reproduces detail. An imaging
system acts on spatial frequencies as a linear filter, and its filter is the
\engterm{optical transfer function} (OTF), whose magnitude is the
\engterm{modulation transfer function} (MTF). For incoherent illumination, the
case of a camera or a microscope viewing a self-luminous or diffusely lit
object, the system is linear in intensity and the OTF is the normalised
autocorrelation of the pupil function. For a circular pupil this gives the
diffraction-limited MTF
\[
\mathrm{MTF}(s) = \frac{2}{\pi}\left[\arccos s - s\sqrt{1 - s^2}\right],
\qquad 0 \le s \le 1,
\]
with $s$ the spatial frequency normalised to the incoherent cutoff
$f_c = 2\,\mathrm{NA}/\lambda$ (equivalently $1/(\lambda\,f/\#)$ for a lens). The
MTF starts at unity at zero frequency, falls smoothly, and reaches zero at the
cutoff, the gentle roll-off that every lens-test chart shows. For coherent
illumination the system is linear in field rather than intensity, and the
transfer function is the pupil itself: a hard pass-band that holds at unit
amplitude out to half the incoherent cutoff and then stops. Figure
\ref{fig:vol04ch11:mtf-cutoff} plots both. The lesson is the one Abbe's mastery
box stated in the spatial domain, read now in the frequency domain: the
numerical aperture and the wavelength fix a hard cutoff frequency, finer detail
diffracts past the pupil rim and is lost, and the only classical handles are a
larger NA or a shorter wavelength.

\input{volumes/04-energy/ch11-optics/figures/fig-mtf-cutoff.tex}

\paragraph{Worked example: MTF and the contrast of a bar target.} An
$f/4$ lens at $\lambda = 550\,\si{\nano\meter}$ has an incoherent cutoff
$f_c = 1/(\lambda\,f/\#) = 1/(550\times 10^{-9}\times 4) = 455\,\text{cycles}/
\si{\milli\meter}$. A test target of $114\,\text{cycles}/\si{\milli\meter}$,
a quarter of the cutoff ($s = 0.25$), is reproduced at a contrast
$\mathrm{MTF}(0.25) = (2/\pi)[\arccos 0.25 - 0.25\sqrt{1 - 0.0625}] =
(2/\pi)[1.318 - 0.242] = 0.685$, so a target imaged at full contrast comes off
the sensor at $69\,\%$ contrast from diffraction alone, before any aberration or
sensor blur. At half the cutoff ($s = 0.5$, $228\,\text{cycles}/\si{\milli
\meter}$) the MTF is $0.391$, and at $80\,\%$ of cutoff it has fallen to $0.10$,
near the threshold a detector can pull out of noise. The number an optical-design
review cares about is the MTF at the detector's Nyquist frequency: a sensor with
$4\,\si{\micro\meter}$ pixels Nyquist-samples at
$1/(2\times 4\times 10^{-6}) = 125\,\text{cycles}/\si{\milli\meter}$, which is
$s = 0.27$ for this lens, where the diffraction MTF is about $0.66$, so the lens
out-resolves the sensor and the system is sensor-limited at $f/4$. Stopping down
to $f/16$ drops the cutoff to $114\,\text{cycles}/\si{\milli\meter}$, below the
sensor Nyquist, and the system becomes diffraction-limited, the trade the camera
worked example traced in the spatial domain and the MTF states in the frequency
domain.

\begin{mastery}
\textbf{Why coherent imaging cuts off at half the frequency yet looks sharper.}
The factor of two between the coherent and incoherent cutoffs is a statement
about what each system is linear in. A coherent system superposes complex field
amplitudes, so its transfer function is the pupil, a band that extends to
$\mathrm{NA}/\lambda$ in field. An incoherent system superposes intensities,
which are squared fields, and squaring in the space domain is autocorrelation in
the frequency domain, so the incoherent transfer function is the pupil
autocorrelated with itself, a band twice as wide, $2\,\mathrm{NA}/\lambda$. The
incoherent system therefore passes finer periodic detail, while the coherent
system passes its narrower band at full, flat amplitude with no roll-off. The
flat band is why coherent images of edges show overshoot and ringing (the sharp
spectral cutoff is a sinc in space, with its side lobes), and the wider
incoherent band is why a microscope run in incoherent (kohler) illumination
resolves finer gratings than the same objective run coherently. The practical
upshot is that the headline resolution number, $0.61\lambda/\mathrm{NA}$, is the
incoherent one, and a coherent instrument such as a scanning laser system must be
quoted against its own narrower cutoff or it will be specified a factor of two
too optimistic.
\end{mastery}

\section{Vision correction: the eye as a variable lens}\pathtag{standard}

The most-manufactured optical component on the planet is the spectacle lens, and
its design is the thin-lens and aberration material of this chapter applied to a
biological imaging system the wearer cannot replace. The relaxed human eye is a
single positive lens of about $+60\,\text{D}$ (dioptres, the reciprocal focal
length in metres, $1\,\text{D} = 1\,\si{\per\meter}$) that should focus a distant
object onto the retina about $22\,\si{\milli\meter}$ behind the cornea. The
cornea supplies roughly two-thirds of the power and the crystalline lens the
rest, with the lens adding power on demand by changing shape, the process of
\engterm{accommodation} that brings near objects into focus.

\engterm{Myopia} (near-sightedness) is the condition of an eye whose optical
power is too high for its axial length, so a distant object focuses in front of
the retina and only near objects reach the retina sharply. The correction is a
diverging spectacle lens that subtracts the excess power. The prescription is
quoted directly in dioptres of the correcting lens: an eye that focuses a distant
object $1.0\,\si{\milli\meter}$ in front of the retina needs about
$-3\,\text{D}$, a common prescription. The far point of an uncorrected myope, the
farthest distance it can focus, is the reciprocal of the excess power; a
$-2.5\,\text{D}$ myope has a far point at $1/2.5 = 0.40\,\si{\meter}$ and cannot
read a sign beyond arm's length without help. The correcting lens forms a virtual
image of a distant object at the eye's far point, which the relaxed eye can then
focus.

\paragraph{Worked example: a myopic prescription and its lens.} A patient cannot
focus beyond $0.40\,\si{\meter}$. The correcting lens must image an object at
infinity to a virtual image at the far point, $s_i = -0.40\,\si{\meter}$ (virtual,
on the incoming side) for $s_o = \infty$, so $1/f = 1/s_i = -2.5\,\text{D}$. The
spectacle prescription is $-2.5\,\text{D}$. If the same patient also reads at
$0.25\,\si{\meter}$, the corrected eye sees that object through the
$-2.5\,\text{D}$ lens. The vergence arithmetic, which adds reciprocal distances
in dioptres, is the quickest route: an object at $0.25\,\si{\meter}$ presents an
incident vergence $U = -1/0.25 = -4\,\text{D}$, the lens of power
$P = -2.5\,\text{D}$ leaves an image vergence $V = P + U = -6.5\,\text{D}$, so
the virtual image sits at $1/6.5 = 0.154\,\si{\meter}$, which the eye must then
accommodate to. A young eye with $3\,\text{D}$ of accommodation handles it; a
presbyopic eye that has lost accommodation needs a second prescription for
reading, the origin of bifocals.

\engterm{Hyperopia} (far-sightedness) is the opposite imbalance, too little power
for the axial length, corrected with a converging lens of positive dioptres.
\engterm{Presbyopia} is the age-related loss of accommodation as the crystalline
lens stiffens; by the sixth decade the lens adds almost no power on demand, and a
near-vision correction becomes necessary regardless of the distance
prescription, delivered as a separate reading lens, a bifocal, or a
progressive-addition lens whose power grades continuously from the distance zone
at the top to the reading zone at the bottom. \engterm{Astigmatism} of the eye,
a cornea curved more steeply along one meridian than another, focuses two
perpendicular line images at different depths and is corrected by a cylindrical
lens component in the prescription, specified by a cylinder power and an axis
angle alongside the spherical power.

The aberration material of this chapter sets the limits of the correction. A
spectacle lens sits $12$ to $15\,\si{\milli\meter}$ in front of the cornea, so a
strong prescription introduces its own off-axis aberrations, chiefly oblique
astigmatism and distortion, which is why a high-minus lens shows a swim of
distortion at the field edge and why the lens is bent into a meniscus and figured
aspheric to push those aberrations down. A contact lens, sitting on the cornea,
removes the vertex-distance problem and the off-axis aberrations almost entirely,
the optical reason a strong prescription is more comfortable as contacts than as
glasses. Corneal refractive surgery reshapes the cornea itself to change its
power, moving the correction from an external lens to the eye's own first surface,
and its failure modes are the chapter's aberrations written into living tissue:
an over-corrected or decentred ablation leaves spherical aberration and coma that
no external lens fully recovers.

\section{Optical fibres}\pathtag{standard}

An \engterm{optical fibre} is a dielectric waveguide consisting of a
high-index core surrounded by a lower-index cladding, with light
confined to the core by total internal reflection (Figure
\ref{fig:vol04ch11:fiber-tir}). The two
predominant fibre types are single-mode (core diameter $\sim 9\,\si{
\micro\meter}$, supporting only the fundamental $\text{LP}_{01}$
mode at $1.55\,\si{\micro\meter}$) and multimode (core diameter
$50\,$or$\,62.5\,\si{\micro\meter}$, supporting many modes).

\input{volumes/04-energy/ch11-optics/figures/fig-fiber-tir.tex}

The fibre's \engterm{numerical aperture} is determined by the
refractive-index step:
\[
\text{NA} = \sqrt{n_{\text{core}}^{2} - n_{\text{clad}}^{2}}.
\]
For a typical telecom fibre with $n_{\text{core}} = 1.467$,
$n_{\text{clad}} = 1.460$, $\text{NA} \approx 0.14$, half-angle
$\sim 8^{\circ}$. The routine \texttt{code/fiber\_na.py} returns the
numerical aperture, the critical angle, and the acceptance angle for any
index pair.

The fibre's attenuation has three components:
\begin{itemize}[leftmargin=*]
  \item Rayleigh scattering, dominant at short wavelengths, $\sim
    1/\lambda^{4}$. At $1.55\,\si{\micro\meter}$ it contributes
    $0.16\,\si{\decibel\per\kilo\meter}$.
  \item Hydroxide-ion absorption peaks at $1.39\,\si{\micro\meter}$
    and $1.24\,\si{\micro\meter}$. Modern fibres are dry (low OH$^{-}$
    contamination) and the peaks are largely suppressed.
  \item Infrared lattice absorption rises above $1.6\,\si{\micro
    \meter}$, setting the long-wavelength limit of silica fibres.
\end{itemize}
The window of minimum attenuation is at $1.55\,\si{\micro\meter}$,
about $0.18\,\si{\decibel\per\kilo\meter}$ in modern single-mode
fibres; this is why the long-haul telecom wavelength is
$1.55\,\si{\micro\meter}$ and why erbium-doped fibre amplifiers
(EDFAs) at the same wavelength are the standard amplifier in the
transmission chain. The measured attenuation against wavelength,
showing the residual hydroxide peak and the $1.55\,\si{\micro\meter}$
minimum, is tabulated in \texttt{data/fiber\_attenuation.csv}.

Dispersion in fibres has three components: material (the
wavelength-dependent refractive index of silica), waveguide (the
mode-dependent phase velocity), and modal (in multimode fibre, the
different propagation times of different modes). Material and
waveguide dispersion cancel at the \engterm{zero-dispersion
wavelength} of $\sim 1.31\,\si{\micro\meter}$ in standard single-
mode fibres; this is why $1.31\,\si{\micro\meter}$ is the second
working telecom window. Dispersion-shifted fibres move the zero-
dispersion wavelength to $1.55\,\si{\micro\meter}$ for combined
low-loss and low-dispersion operation.

The standard single-mode fibre deployed across most of the installed base
carries its zero-dispersion wavelength at $1.31\,\si{\micro\meter}$ and a
chromatic dispersion near $17\,\si{\pico\second/(\nano\meter\,\kilo\meter)}$
at $1.55\,\si{\micro\meter}$, so a long-haul system run in the low-loss
$1.55\,\si{\micro\meter}$ window pays for that loss saving with a dispersion
penalty that grows linearly with distance. The two engineering answers are
dispersion-compensating modules, which insert a length of fibre with large
opposite-sign dispersion to undo the accumulated spread, and dispersion-
managed line design, which alternates spans of positive and negative
dispersion so the pulse periodically recompresses. The end-to-end case study
below carries this trade through to a number. Optical amplification changed
the economics of the whole chain: before the erbium-doped fibre amplifier,
every span had to terminate at an electronic regenerator that detected,
retimed, and retransmitted the signal, one per wavelength; the EDFA amplifies
many wavelengths in one device, in the optical domain, with no per-channel
electronics, which is what made dense wavelength-division multiplexing and the
modern long-haul internet economical.

\section{Worked examples}\pathtag{standard}

Three examples carry the chapter's formulas through to numbers an
engineer would defend.

\paragraph{A camera lens and its diffraction floor.} A
$50\,\si{\milli\meter}$ lens stopped to $f/16$ has an entrance pupil of
$50/16 = 3.1\,\si{\milli\meter}$. The Airy-disc radius at the sensor is
$1.22\lambda(f/\#) = 1.22\times 550\times10^{-9}\times 16 = 10.7\,\si{
\micro\meter}$. A full-frame sensor with $4\,\si{\micro\meter}$ pixels
spreads each point over a five-pixel blur at $f/16$, so stopping down
past about $f/11$ buys depth of field at the cost of resolution. The
optimum aperture for a given sensor balances the growing diffraction
blur against the shrinking aberration blur, and it sits two to three
stops down from wide open for most lenses.

\paragraph{A grating spectrometer's reach.} A grating ruled at
$600\,\si{\per\milli\meter}$ has period $d = 1/600\,\si{\milli\meter} =
1.67\,\si{\micro\meter}$. At normal incidence the first-order angle for
green light at $550\,\si{\nano\meter}$ is
$\theta_{1} = \arcsin(\lambda/d) = \arcsin(0.330) = 19.3^{\circ}$, and
the highest order is $|m| \le d/\lambda = 3.0$, so three orders reach
each side. The angular dispersion $d\theta/d\lambda = m/(d\cos\theta)$
grows with order, which is why high-resolution spectrographs work in
high orders. The values match \texttt{code/diffraction.py} run on
$d = 1.667\,\si{\micro\meter}$.

\paragraph{A fibre link's power budget.} A single-mode link launches
$0\,\si{\decibel}$-referenced power of $1\,\si{\milli\watt}$
($0\,\text{dBm}$) into $80\,\si{\kilo\meter}$ of fibre at
$0.20\,\si{\decibel\per\kilo\meter}$, a span loss of $16\,\si{
\decibel}$. With two connector pairs at $0.5\,\si{\decibel}$ each and a
$3\,\si{\decibel}$ margin, the received power is
$0 - 16 - 1 - 3 = -20\,\text{dBm}$, comfortably above a typical
receiver sensitivity of $-28\,\text{dBm}$ at $10\,\si{\giga\bit\per
\second}$. No mid-span amplifier is needed at this distance; the budget
shows the headroom that lets the same fibre run to $120\,\si{\kilo
\meter}$ before an EDFA becomes necessary.

\paragraph{A single-layer anti-reflection coating.} A camera lens of
crown glass ($n_{\text{sub}} = 1.52$) loses about $4.3\,\%$ of the light at
each air-glass surface, computed from the normal-incidence Fresnel
reflectance $R = [(n_{\text{sub}}-1)/(n_{\text{sub}}+1)]^{2} =
(0.52/2.52)^{2} = 0.043$. A quarter-wave anti-reflection layer cancels the
reflection at one wavelength when its index is the geometric mean of the
media it bridges, $n_{\text{film}} = \sqrt{n_{\text{air}}n_{\text{sub}}} =
\sqrt{1.52} = 1.23$. No durable solid has so low an index; the practical
choice is magnesium fluoride at $n = 1.38$, which leaves a residual
reflectance $R = [(n_{\text{air}}n_{\text{sub}} - n_{\text{film}}^{2})/
(n_{\text{air}}n_{\text{sub}} + n_{\text{film}}^{2})]^{2} = [(1.52 - 1.904)/
(1.52 + 1.904)]^{2} = (0.384/3.424)^{2} = 0.0126$, about $1.3\,\%$ at the
design wavelength rather than the ideal zero. The layer thickness is a
quarter-wave in the film, $t = \lambda/(4 n_{\text{film}}) = 550\times
10^{-9}/(4\times 1.38) = 99.6\,\si{\nano\meter}$ for a green design point.
Figure \ref{fig:vol04ch11:ar-coating} plots the band shape: a single layer
gives a broad shallow minimum, a two-layer V-coat a deep narrow null. A
six-element lens with twelve air-glass surfaces would lose
$1 - 0.957^{12} = 41\,\%$ of the light uncoated; single-layer coating drops
that to $1 - 0.987^{12} = 15\,\%$, and a multilayer broadband coat below
$3\,\%$, which is why every photographic lens is coated.

\input{volumes/04-energy/ch11-optics/figures/fig-ar-coating.tex}

\paragraph{A Fabry-Perot etalon and its finesse.} Two parallel partially
reflecting mirrors of reflectance $R$ separated by a gap $d$ transmit light
only at wavelengths for which the round-trip path is an integer number of
wavelengths, $2 n d\cos\theta = m\lambda$. The transmission is the Airy
function, whose peaks are sharpened as the mirror reflectance rises. The
\engterm{finesse}
\[
\mathcal{F} = \frac{\pi\sqrt{R}}{1 - R}
\]
measures the ratio of the free spectral range to the width of a transmission
peak. A pair of mirrors at $R = 0.95$ has $\mathcal{F} = \pi\sqrt{0.95}/0.05
= 61$. The free spectral range, the frequency gap between adjacent
transmission orders, is $\Delta f_{\text{FSR}} = c/(2 n d)$; for an air gap
$d = 5\,\si{\milli\meter}$ it is $3\times10^{8}/(2\times 0.005) =
30\,\si{\giga\hertz}$. The resolvable linewidth is the free spectral range
divided by the finesse, $30\,\si{\giga\hertz}/61 = 0.49\,\si{\giga\hertz}$,
so this etalon resolves spectral features half a gigahertz apart, fine
enough to separate the longitudinal modes of an unstabilised gas laser. The
etalon is the working element of laser line-narrowing, optical-spectrum
analysers, and the wavelength-locking filters of dense wavelength-division
multiplexing.

\paragraph{A Gaussian beam's waist and divergence.} A real laser beam is not
a ray; it is a Gaussian-profile beam whose radius follows
$w(z) = w_{0}\sqrt{1 + (z/z_{R})^{2}}$, with the waist $w_{0}$ the minimum
radius and the Rayleigh range $z_{R} = \pi w_{0}^{2}/\lambda$ the distance
over which the beam stays near-collimated. A helium-neon beam at
$\lambda = 633\,\si{\nano\meter}$ focused to a waist
$w_{0} = 0.5\,\si{\milli\meter}$ has $z_{R} = \pi(5\times10^{-4})^{2}/
(633\times10^{-9}) = 1.24\,\si{\meter}$ and a far-field divergence half-angle
$\theta = \lambda/(\pi w_{0}) = 633\times10^{-9}/(\pi\times 5\times10^{-4}) =
4.0\times10^{-4}\,\si{\radian}$, about $0.4\,\si{\milli\radian}$. Over a
$100\,\si{\meter}$ throw the beam grows to
$w \approx \theta z = 4.0\times10^{-4}\times 100 = 4\,\si{\centi\meter}$
radius. Focusing the same beam to a tighter $w_{0} = 25\,\si{\micro\meter}$
spot, as a laser cutter demands, shrinks the Rayleigh range to
$z_{R} = \pi(25\times10^{-6})^{2}/(633\times10^{-9}) = 3.1\,\si{\milli
\meter}$, so the cut stays in focus over only a few millimetres of depth.
Figure \ref{fig:vol04ch11:gaussian-beam} shows the hyperbolic envelope and
its far-field cone. The waist and divergence trade against each other through
their product $w_{0}\theta = \lambda/\pi$, a beam-quality invariant the
diffraction limit fixes and that no lens can beat.

\input{volumes/04-energy/ch11-optics/figures/fig-gaussian-beam.tex}

\paragraph{Hyperfocal distance and the depth-of-field budget.} A photographer
who must hold a scene sharp from the foreground to the horizon sets the lens to
its \engterm{hyperfocal distance}, the focus setting whose far depth-of-field
limit reaches infinity. The hyperfocal distance is
$H = f^{2}/(N c) + f$, where $f$ is the focal length, $N$ the relative aperture
($f/\#$), and $c$ the circle of confusion, the largest blur diameter the
final image tolerates. For a $35\,\si{\milli\meter}$ lens at $f/8$ on a
full-frame sensor with $c = 30\,\si{\micro\meter}$,
\[
H = \frac{(0.035)^{2}}{8\times 30\times 10^{-6}} + 0.035
= \frac{1.225\times 10^{-3}}{2.4\times 10^{-4}} + 0.035
= 5.1 + 0.035 \approx 5.1\,\si{\meter}.
\]
Focusing at $H$ holds everything from $H/2 = 2.6\,\si{\meter}$ to infinity in
acceptable focus. Stopping down one stop to $f/11$ shrinks $H$ to about
$3.7\,\si{\meter}$ and pulls the near limit in to $1.85\,\si{\meter}$, at the
cost of the diffraction blur that the camera worked example traced. The two
effects pull against each other: opening the aperture sharpens the plane of
focus by cutting diffraction and shrinks the depth that stays sharp, while
closing it deepens the sharp zone and softens every plane to diffraction. The
working sweet spot for landscape work sits two to three stops down from wide
open, where the depth of field is generous and the diffraction blur has not yet
overtaken the sensor's pixel pitch. The circle of confusion is the editorial
choice in the budget; a print viewed close demands a smaller $c$, which pushes
$H$ out and the depth of field in.

\paragraph{A two-lens system and its back focal distance.} Two thin lenses of
focal lengths $f_{1}$ and $f_{2}$ separated by a distance $d$ combine into a
system whose effective focal length follows the Gullstrand equation,
\[
\frac{1}{f_{\text{sys}}} = \frac{1}{f_{1}} + \frac{1}{f_{2}}
- \frac{d}{f_{1} f_{2}}.
\]
Take a $+100\,\si{\milli\meter}$ objective and a $-50\,\si{\milli\meter}$
diverging element spaced $d = 60\,\si{\milli\meter}$ apart, the layout of a
simple telephoto. Then
\[
\frac{1}{f_{\text{sys}}} = \frac{1}{100} + \frac{1}{-50}
- \frac{60}{100\times(-50)}
= 0.01 - 0.02 + 0.012 = 0.002\,\si{\per\milli\meter},
\]
so $f_{\text{sys}} = 500\,\si{\milli\meter}$. The system has a long focal
length packed into a short barrel, the defining trick of the telephoto. The
back focal distance, measured from the rear lens to the focal plane, is
\[
\text{BFD} = f_{\text{sys}}\left(1 - \frac{d}{f_{1}}\right)
= 500\left(1 - \frac{60}{100}\right) = 200\,\si{\milli\meter}.
\]
The physical length from the front element to the focal plane is
$d + \text{BFD} = 260\,\si{\milli\meter}$, about half the $500\,\si{\milli
\meter}$ a single lens of the same focal length would demand, which is why the
diverging rear group earns its place. The same arithmetic with a converging
rear group and the lenses spaced near $f_{1} + f_{2}$ builds a beam expander
or a relay, so the Gullstrand equation is the first tool of any multi-element
layout.

\paragraph{A polariser stack and Malus's law.} Linearly polarised light of
intensity $I_{0}$ passing through an analyser whose transmission axis is at
angle $\theta$ to the polarisation emerges with intensity
$I = I_{0}\cos^{2}\theta$, Malus's law. Two crossed polarisers ($\theta =
90^{\circ}$) pass nothing, $\cos^{2}90^{\circ} = 0$. Insert a third sheet at
$\phi = 45^{\circ}$ between them and the transmitted intensity becomes the
product of two Malus factors,
\[
\frac{I}{I_{0}} = \cos^{2}(45^{\circ})\,\cos^{2}(90^{\circ}-45^{\circ})
= \frac{1}{2}\times\frac{1}{2} = \frac{1}{4}.
\]
A pair that blocked all the light now passes a quarter of it, because each
sheet projects the field onto its own axis and the projection discards the
component the previous sheet enforced. Figure \ref{fig:vol04ch11:malus-stack}
plots the single-analyser cosine-squared law and the inserted-sheet
$\tfrac{1}{4}\sin^{2}2\phi$ curve that peaks at $45^{\circ}$. The general
result for $N$ sheets each rotated by $90^{\circ}/N$ from the last is
$I/I_{0} = \cos^{2N}(90^{\circ}/N)$, which rises toward unity as $N$ grows: a
stack of many slightly-rotated polarisers can rotate the plane of polarisation
through a right angle while passing nearly all the light, the quantum-Zeno-like
limit that explains the brightness of a liquid-crystal display. Each display
pixel is a voltage-controlled birefringent layer between crossed polarisers,
switching between the dark crossed state and a bright rotated state.

\input{volumes/04-energy/ch11-optics/figures/fig-malus-stack.tex}

\paragraph{Radiation pressure and an optical trap.} Light carries momentum, so
a beam exerts a force on whatever absorbs or redirects it. A beam of power $P$
fully absorbed delivers a force $F = P/c$; fully reflected, it delivers
$F = 2P/c$ because the photon momentum reverses. A $10\,\si{\watt}$ laser
fully reflected from a mirror pushes with
\[
F = \frac{2P}{c} = \frac{2\times 10}{3\times 10^{8}}
= 6.7\times 10^{-8}\,\si{\newton},
\]
about seventy nanonewtons, negligible against gravity on any laboratory mass
but decisive on a micron-scale particle. An \engterm{optical trap} (optical
tweezers) exploits the gradient of a tightly focused beam: a transparent
microsphere of higher index than its surroundings is drawn toward the intensity
maximum because refracting the converging rays bends their momentum outward, and
by reaction the bead is pulled inward toward the focus. The trap stiffness for a
$1\,\si{\micro\meter}$ polystyrene bead in a $10\,\si{\milli\watt}$
near-infrared beam is of order
$\kappa \sim 0.1\,\si{\pico\newton}/\si{\nano\meter}$, so a
$10\,\si{\nano\meter}$ displacement restores with a force near
$1\,\si{\pico\newton}$, the scale of single-molecule motor forces. The trap is
the working instrument behind the measurement of kinesin step forces and the
mechanical unfolding of single proteins, and its force scale follows directly
from the photon-momentum bookkeeping that the radiation-pressure formula sets.

\paragraph{A grating's resolving power and a sodium doublet.} A spectrometer's
ability to separate two close wavelengths is its \engterm{resolving power}
$\mathcal{R} = \lambda/\Delta\lambda$, and for a diffraction grating it equals
the order times the number of illuminated grooves, $\mathcal{R} = m N$. A
grating ruled at $1200\,\si{\per\milli\meter}$ over an illuminated width of
$25\,\si{\milli\meter}$ presents $N = 1200\times 25 = 30\,000$ grooves. In
first order ($m = 1$) its resolving power is $\mathcal{R} = 30\,000$, so at the
sodium D lines near $589\,\si{\nano\meter}$ it resolves
$\Delta\lambda = \lambda/\mathcal{R} = 589/30\,000 = 0.0196\,\si{\nano\meter}$.
The sodium doublet is split by $0.597\,\si{\nano\meter}$ (the D$_{1}$
line at $589.59\,\si{\nano\meter}$ and D$_{2}$ at $588.99\,\si{\nano\meter}$), so
this grating separates the pair with a thirty-fold margin in first order, and
would do so even in a beam clipped to a third of its width. The formula carries
a lesson: resolving power buys nothing from the groove density alone; it comes
from the total number of grooves the beam actually covers, which is why a
high-resolution spectrograph wants a large, fully illuminated grating rather
than merely a finely ruled one. Working in a higher order $m$ raises
$\mathcal{R}$ proportionally but shrinks the free spectral range $\lambda/m$ in
step, so an echelle spectrograph runs in high order and uses a cross-disperser
to unstack the overlapping orders.

\paragraph{A wedge film and a flatness gauge.} Pressing an optical flat against
a polished surface and illuminating it from above with monochromatic light
turns the thin air wedge between them into a contour map of the gap. The two
interfering beams reflect from the bottom of the flat and the top of the test
surface, separated by twice the local air gap $t$, and a reflection at the
lower (air-to-glass) surface adds the $\lambda/2$ phase step the thin-film
arithmetic tracks. Dark fringes fall where $2t = m\lambda$, so adjacent dark
fringes mark a gap change of $\lambda/2$. Under a HeNe source at
$\lambda = 633\,\si{\nano\meter}$ each fringe is a contour of
$\lambda/2 = 316\,\si{\nano\meter}$ of height. A surface that shows ten
straight, evenly spaced fringes across a $25\,\si{\milli\meter}$ aperture
departs from flat by $10\times 316\,\si{\nano\meter} = 3.2\,\si{\micro\meter}$
across that width, a slope, while fringes that curve reveal a convex or
concave figure and fringes that close into rings reveal a spherical error. A
shop specifying flatness ``to a quarter wave'' is asking that the fringe
departure from straight stay under half a fringe across the part, since a
quarter-wave of surface height is a half-wave of optical path and so half a
fringe. The same wedge, filled with an unknown transparent film instead of air,
measures the film's optical thickness $n t$ from the fringe count, the working
principle of a Fizeau interferometer.

\paragraph{A prism at minimum deviation.} A glass prism disperses light because
its refractive index falls with wavelength, and the cleanest way to measure
that index is to find the prism's \engterm{minimum deviation} angle. A ray
passing symmetrically through a prism of apex angle $A$ suffers the least total
deviation $\delta_{\min}$, and at that symmetric pass the index follows the
prism formula
\[
n = \frac{\sin\!\big[(A + \delta_{\min})/2\big]}{\sin(A/2)}.
\]
A $60^{\circ}$ equilateral prism that deviates green light by a measured
$\delta_{\min} = 38.9^{\circ}$ has
\[
n = \frac{\sin\!\big[(60^{\circ} + 38.9^{\circ})/2\big]}{\sin(30^{\circ})}
= \frac{\sin(49.45^{\circ})}{0.5} = \frac{0.7597}{0.5} = 1.519,
\]
the index of a typical crown glass at $550\,\si{\nano\meter}$. Repeating the
measurement at the blue F line and the red C line gives the two indices whose
difference fixes the glass's dispersion, quoted as the Abbe number
$V_{d} = (n_{d}-1)/(n_{F}-n_{C})$ that the achromat design of the aberration
section turns on; a crown at $V_{d}\approx 60$ disperses gently, a dense flint
at $V_{d}\approx 30$ strongly. The minimum-deviation method needs only a
spectrometer table and an angle, with no contact with the glass, which is why
it remained the reference index measurement for a century.

\paragraph{The atmosphere's Fried parameter and an adaptive-optics budget.} The
seeing that caps a ground telescope has a length scale, the \engterm{Fried
parameter} $r_{0}$, the aperture over which the atmospheric wavefront stays
flat to about a radian. A telescope of diameter $D \le r_{0}$ is
diffraction-limited; one with $D \gg r_{0}$ delivers seeing-limited resolution
$\sim\lambda/r_{0}$ no matter how large its mirror. At a good site $r_{0}$ is
about $15\,\si{\centi\meter}$ at $550\,\si{\nano\meter}$, which gives a seeing
disc of $\lambda/r_{0} = 550\times10^{-9}/0.15 = 3.7\,\si{\micro\radian}$, or
$0.76$ arc-second, the familiar number. An adaptive-optics system flattens the
wavefront by tiling the aperture with deformable-mirror actuators spaced about
$r_{0}$ apart, so a $4\,\si{\meter}$ telescope needs of order
$(D/r_{0})^{2} = (4/0.15)^{2} \approx 710$ actuators to correct in the visible.
The correction must keep pace with the atmosphere, which evolves on the
timescale $\tau_{0} \approx r_{0}/v$ for a wind speed $v$; at
$v = 10\,\si{\meter\per\second}$ this is $0.15/10 = 15\,\si{\milli\second}$, so
the loop must run at $\sim 1\,\si{\kilo\hertz}$ to stay ahead of it. Both
demands ease toward the infrared, where $r_{0}$ grows as $\lambda^{6/5}$:
at $1.6\,\si{\micro\meter}$ the Fried parameter is about
$0.15\times(1600/550)^{6/5} = 0.55\,\si{\meter}$, the actuator count drops to
$(4/0.55)^{2}\approx 53$, and the timescale lengthens, which is why the first
working adaptive optics ran in the near-infrared and the visible came later.

\begin{mastery}
\textbf{The Abbe diffraction limit as a spatial-frequency cutoff.} The
resolution formula $d_{\min} = 0.61\lambda/\text{NA}$ is better read as
a statement about spatial frequencies than about disc radii. An object
illuminated by a wave carries information in its spatial Fourier
components; a component of period $\Lambda$ diffracts light at angle
$\sin\theta = \lambda/\Lambda$. The objective collects only the
diffracted orders that fall within its acceptance cone, $\sin\theta \le
\text{NA}$, so it passes only spatial frequencies up to
$1/\Lambda_{\min} = \text{NA}/\lambda$, a hard low-pass filter in the
spatial-frequency domain. Finer detail diffracts past the rim of the
objective and never enters the image. This is Abbe's 1873 insight, and
it explains why oil immersion (raising $n$, hence NA above unity) and
shorter wavelengths are the only classical routes to finer resolution,
and why the super-resolution methods of the imaging section must break
the assumptions (linear response, far-field collection) on which the
cutoff rests.
\end{mastery}

\begin{mastery}
\textbf{Temporal and spatial coherence as two independent budgets.}
Coherence is two separate properties, and an instrument can have plenty of
one and none of the other. \emph{Temporal} coherence is the predictability of
the phase along the direction of propagation; it is set by the source's
spectral width through $L_{c} = c/\Delta f$ and limits the path-length
imbalance over which interference fringes survive. A white-light source with
$\Delta f \sim 3\times10^{14}\,\si{\hertz}$ has $L_{c}$ of order a wavelength,
so its fringes appear only at near-zero path difference, the property that
white-light interferometry exploits to find an absolute zero-path point.
\emph{Spatial} coherence is the predictability of the phase across the
wavefront, transverse to propagation; it is set by the source's angular size
seen from the observation point. The transverse coherence width is
$d_{c} \approx \lambda/\Delta\phi$, where $\Delta\phi$ is the angular diameter
the source subtends. Sunlight, with the Sun's $0.5^{\circ}$ angular diameter
($\Delta\phi \approx 9\times10^{-3}\,\si{\radian}$), has a spatial coherence
width of only $d_{c} \approx 550\times10^{-9}/(9\times10^{-3}) = 60\,\si{
\micro\meter}$ at the ground, which is why a double-slit experiment in raw
sunlight needs slit separations below about $60\,\si{\micro\meter}$ to show
fringes. The same arithmetic, run in reverse, is stellar interferometry: the
Michelson stellar interferometer measures a star's angular diameter by
finding the baseline at which its fringes wash out, turning a spatial-
coherence measurement into an angular size. A laser is the rare source that is
long in both budgets at once; a pinhole-filtered thermal source can be made
spatially coherent while staying temporally incoherent, and a narrowband-
filtered extended source the reverse.
\end{mastery}

\section{An end-to-end case study: a metro fibre link}\pathtag{standard}

The chapter's wave optics, fibre physics, and budgeting arithmetic pay off
when an engineer must clear a real communication link for service from the
component data sheets alone. We carry a metropolitan-area fibre link from the
transmitter laser through coupling, fibre attenuation over distance,
connector and splice losses, an optical amplifier, chromatic and modal
dispersion, the bandwidth-distance limit, the receiver sensitivity, and two
margins, ending in a verdict an engineer would sign. The link is a good
vehicle because every fibre move of the chapter appears in it once: launch
power, attenuation, numerical-aperture coupling, dispersion, and the
signal-to-noise reasoning behind receiver sensitivity.

\subsection{The link and its components}

A carrier must connect two data centres $120\,\si{\kilo\meter}$ apart over a
single-mode fibre at $1.55\,\si{\micro\meter}$, carrying
$10\,\si{\giga\bit\per\second}$ on one wavelength. The transmitter is a
distributed-feedback laser diode launching $+3\,\text{dBm}$
($2\,\si{\milli\watt}$) with a spectral linewidth of
$0.05\,\si{\nano\meter}$ under direct modulation. The fibre is standard
single-mode fibre with attenuation $0.20\,\si{\decibel\per\kilo\meter}$ at
$1.55\,\si{\micro\meter}$ and chromatic dispersion
$D = 17\,\si{\pico\second/(\nano\meter\,\kilo\meter)}$. The route has two
connector pairs at the data-centre ends ($0.5\,\si{\decibel}$ each) and six
fusion splices along the buried cable ($0.1\,\si{\decibel}$ each). The
receiver is an avalanche-photodiode module with sensitivity
$-24\,\text{dBm}$ at $10\,\si{\giga\bit\per\second}$ for a bit-error rate of
$10^{-12}$.

\subsection{Coupling and the power budget}

The laser launches $+3\,\text{dBm}$, but coupling the diode's output into the
fibre's $9\,\si{\micro\meter}$ mode field carries a loss; a pigtailed
distributed-feedback laser couples with about $1\,\si{\decibel}$ of insertion
loss, so $+2\,\text{dBm}$ enters the fibre. The span attenuation over
$120\,\si{\kilo\meter}$ is
\[
A_{\text{fibre}} = 0.20\,\si{\decibel\per\kilo\meter}\times 120\,\si{\kilo
\meter} = 24\,\si{\decibel}.
\]
Connectors and splices add
$2\times 0.5 + 6\times 0.1 = 1.6\,\si{\decibel}$. Without amplification the
received power would be $+2 - 24 - 1.6 = -23.6\,\text{dBm}$, barely above the
$-24\,\text{dBm}$ sensitivity, leaving almost no margin. Practice
requires a margin against fibre ageing, repair splices after a future cable
cut, and connector contamination; a working design carries
$6\,\si{\decibel}$. The budget is short, so a mid-span erbium-doped fibre
amplifier is placed at the $60\,\si{\kilo\meter}$ point. It restores
$14\,\si{\decibel}$, lifting the received power to about $-18\,\text{dBm}$,
which clears the sensitivity by $6\,\si{\decibel}$. Figure
\ref{fig:vol04ch11:link-budget} carries the power level down the route from
launch to receiver, with the amplifier step and the closing margin marked.

\input{volumes/04-energy/ch11-optics/figures/fig-link-budget.tex}

\subsection{The dispersion budget}

Power margin is necessary but not sufficient; a link can deliver ample power
and still fail because the pulses have smeared into one another. The
\engterm{bandwidth-distance limit} comes from chromatic dispersion, the
wavelength dependence of the fibre's group velocity. A pulse of spectral
width $\Delta\lambda$ spreads in time over a length $L$ by
\[
\Delta\tau = D\,\Delta\lambda\,L = 17\,\si{\pico\second/(\nano\meter\,\kilo
\meter)}\times 0.05\,\si{\nano\meter}\times 120\,\si{\kilo\meter} = 102\,\si{
\pico\second}.
\]
The bit period at $10\,\si{\giga\bit\per\second}$ is
$1/(10\times10^{9}) = 100\,\si{\pico\second}$. The pulse spread of
$102\,\si{\pico\second}$ already exceeds one bit period, so adjacent symbols
overlap and the receiver cannot tell them apart; the link fails on dispersion
even though it passes on power. The working rule keeps the spread under about
a quarter of the bit period, so this route needs dispersion compensation. A
dispersion-compensating fibre module with total dispersion
$-1700\,\si{\pico\second/\nano\meter}$ cancels the
$D L = 17\times 120 = 2040\,\si{\pico\second/\nano\meter}$ of the span down to
a residual $340\,\si{\pico\second/\nano\meter}$, leaving a spread of
$0.05\times 340 = 17\,\si{\pico\second}$, comfortably within the quarter-bit
budget. The compensating module adds its own $3$ to $4\,\si{\decibel}$ of
insertion loss, which the power budget above must absorb; the
$6\,\si{\decibel}$ margin covers it with room to spare, and the amplifier gain
can be trimmed up if it does not.

\subsection{Modal dispersion and the single-mode choice}

The route uses single-mode fibre for a reason this case makes concrete. A
multimode fibre at $1.55\,\si{\micro\meter}$ carries hundreds of modes, each
with a different path length down the core; the resulting modal dispersion in
a step-index multimode fibre is of order
$10\,\si{\nano\second\per\kilo\meter}$, which over $120\,\si{\kilo\meter}$
would smear a pulse by more than a microsecond, ten thousand bit periods. No
amount of power margin recovers that. The single-mode core diameter of
$9\,\si{\micro\meter}$ removes modal dispersion entirely by supporting only
the fundamental mode, which is why long-haul and metro links are single-mode
without exception and multimode survives only in short data-centre runs where
its cheaper transceivers win.

\subsection{Verdict}

The link closes. The power budget delivers $-18\,\text{dBm}$ to a
$-24\,\text{dBm}$ receiver, a $6\,\si{\decibel}$ margin that absorbs the
dispersion-compensating module's insertion loss and the ageing allowance. The
dispersion budget, uncorrected, fails by a hair; with a dispersion-
compensating module sized to the span it passes with a four-fold margin. The
single-mode choice removes modal dispersion as a concern. The design that
ships is a $+3\,\text{dBm}$ distributed-feedback transmitter, a mid-span
erbium-doped fibre amplifier at $60\,\si{\kilo\meter}$, a dispersion-
compensating module at the receiver end, and an avalanche-photodiode
receiver, with both the power and the dispersion budgets cleared and
documented. The verdict illustrates the chapter's claim by scale: a fibre
link is two budgets at once, power and time, and a link that passes one while
failing the other does not work. The engineer who carries only the power
budget ships a link that powers up and cannot pass traffic.

\begin{principle}[A link is two budgets, power and time]
An optical communication link must close a power budget and a dispersion
budget independently. Adequate received power guarantees only that the
detector sees light; it says nothing about whether the pulses remain
distinguishable in time. A link is cleared for service only when both the
power margin and the bandwidth-distance product are documented against the
required bit rate and distance.
\end{principle}

\section{A second end-to-end case study: designing a survey telescope}
\pathtag{standard}

The fibre link carried the wave-optics and budgeting machinery of the chapter
to a number. A telescope carries the geometric-optics, diffraction, and
detector reasoning to a different number, and the two case studies together
span the chapter's two halves. We design a ground-based survey telescope from a
science requirement, work through aperture, focal ratio, plate scale, the
contest between diffraction and atmospheric seeing, light-gathering, and the
signal-to-noise calculation that fixes the limiting magnitude, and close with a
verdict an instrument scientist would sign. Every imaging-section formula
appears once: the Rayleigh limit, the numerical aperture in the guise of focal
ratio, the diffraction-limited spot, and the photon-shot-noise floor the
detector exercises introduced.

\subsection{The requirement and the aperture}

A regional observatory wants a telescope that surveys a wide field nightly to
magnitude $V = 21$ at a signal-to-noise ratio of $5$ in a
$300\,\si{\second}$ exposure, sited where the median seeing is
$1.0$ arc-second. The science sets two quantities at once: the aperture, which
must gather enough light to reach magnitude $21$ in five minutes, and the
plate scale, which must match the camera's pixels to the seeing disc so that no
resolution is thrown away and no pixels are wasted.

Aperture follows from the photon budget. A star of visual magnitude $V = 0$
delivers about $10^{4}$ photons per second per square centimetre per
\AA ngstr\"om above the atmosphere; the broad $V$ band is about
$880\,\text{\AA}$ wide, so a zeroth-magnitude star delivers of order
$8.8\times 10^{6}\,\text{photons}/(\text{s}\cdot\text{cm}^{2})$ at the top of
the atmosphere.
The magnitude scale is logarithmic, each five magnitudes a factor of $100$ in
flux, so a $V = 21$ star is fainter by $100^{21/5} = 10^{8.4} = 2.5\times
10^{8}$ and delivers about
$8.8\times 10^{6}/2.5\times 10^{8} = 0.035\,\text{photons}/(\text{s}\cdot
\text{cm}^{2})$. A
realistic system throughput, the product of atmospheric transmission, mirror
reflectivities, filter bandpass, and detector quantum efficiency, is about
$0.3$, so the detected rate is about $0.011$ photons per second per square
centimetre. The layout of primary, secondary, and folded focus is the
arrangement of Figure \ref{fig:vol04ch11:telescope-layout}.

\input{volumes/04-energy/ch11-optics/figures/fig-telescope-layout.tex}

\subsection{The signal-to-noise calculation and the limiting magnitude}

A faint star competes against the sky background, the read noise of the
detector, and the shot noise of its own photons. The signal-to-noise ratio over
an exposure of duration $t$ collecting a stellar count $S$ over a seeing-sized
aperture of $n_{\text{pix}}$ pixels, each carrying a sky count $B$ and a read
noise $\sigma_{r}$, is
\[
\text{SNR} = \frac{S t}{\sqrt{S t + n_{\text{pix}}\,B t
+ n_{\text{pix}}\,\sigma_{r}^{2}}},
\]
the standard CCD equation. For a faint star against a bright sky the middle
term dominates, and the ratio reduces to the background-limited form
$\text{SNR} \approx S t/\sqrt{n_{\text{pix}} B t}$, which grows as the square
root of both aperture area and exposure time. We require $\text{SNR} = 5$ at
$V = 21$ in $t = 300\,\si{\second}$. The stellar count for an aperture of
diameter $D$ is the detected rate times the collecting area,
$S = 0.011\times \pi (D/2)^{2}$ with $D$ in centimetres. A dark site sky
background at $V \approx 21.5\,\text{mag}/\text{arcsec}^{2}$ contributes, over a
seeing-disc area of about $\pi(0.5)^{2} = 0.79\,\text{arcsec}^{2}$, a count
comparable to the star, so the background-limited estimate needs the area to
reach $\text{SNR} = 5$. Carrying the arithmetic, a clear aperture near
$D = 1.0\,\si{\meter}$ (collecting area $7854\,\si{cm^{2}}$) delivers
$S \approx 86\,\text{photons}/\text{s}$, a $300\,\si{\second}$ stellar count of
about
$2.6\times 10^{4}$, against a sky-plus-read variance that yields
$\text{SNR}\approx 5$ at $V = 21$. The requirement sets the aperture at one
metre. Doubling the aperture to two metres quadruples the light and reaches the
same SNR at $V = 22.5$, one and a half magnitudes deeper, the steep payoff that
drives observatories to the largest aperture the budget allows.

\subsection{Focal ratio, plate scale, and pixel matching}

Aperture sets the light and the diffraction limit; the focal ratio sets the
plate scale, the angle on the sky that maps to a millimetre at the focal plane.
The \engterm{plate scale} in radians per millimetre is $1/f$ with $f$ the focal
length, conventionally quoted in arc-seconds per millimetre,
\[
\text{plate scale} = \frac{206265}{f\,[\si{\milli\meter}]}
\quad\text{arcsec}/\text{mm}.
\]
A one-metre $f/4$ telescope has $f = 4000\,\si{\milli\meter}$ and a plate scale
of $206265/4000 = 52\,\text{arcsec}/\text{mm}$, equivalently
$0.052\,\text{arcsec}/\si{\micro\meter}$. The seeing disc of $1.0$ arc-second then
spans $1.0/0.052 = 19\,\si{\micro\meter}$ at the focal plane. Good sampling
puts about two pixels across the seeing disc (the Nyquist rule that no detail
finer than the resolution element should fall between samples), so the camera
wants pixels near $9$ to $10\,\si{\micro\meter}$, which a scientific CCD
provides. Choosing $f/4$ rather than a slower $f/8$ keeps the plate scale
coarse enough that a modest sensor covers a wide field, the survey
requirement, at the cost of tighter alignment and aberration tolerances that a
fast focal ratio demands. The diffraction limit at this aperture is
$1.22\lambda/D = 1.22\times 550\times 10^{-9}/1.0 = 6.7\times
10^{-7}\,\si{\radian} = 0.14$ arc-second, far below the one-arc-second seeing,
so the telescope is seeing-limited, not diffraction-limited, the universal
condition of a ground-based instrument without adaptive optics. Figure
\ref{fig:vol04ch11:seeing-vs-diffraction} places this design on the contest
between the two limits.

\input{volumes/04-energy/ch11-optics/figures/fig-seeing-vs-diffraction.tex}

\subsection{Verdict}

The design closes against its requirement. A one-metre $f/4$ primary reaches
$V = 21$ at $\text{SNR} = 5$ in a five-minute exposure under one-arc-second
seeing, the science requirement, with a $52\,\text{arcsec}/\text{mm}$ plate
scale that Nyquist-samples the seeing disc onto a CCD of $9\,\si{\micro\meter}$
pixels. The aperture is set by the photon budget, not the diffraction limit,
because the atmosphere wastes the diffraction-limited resolution the optics
could in principle reach; the telescope delivers one-arc-second images and is
worth no more aperture for resolution, only for depth. The instrument that
ships is a one-metre $f/4$ reflector with a wide-field CCD camera matched to
the seeing, a design whose every number traces back to the survey requirement
through the chain the chapter built. The two case studies make the same point
from opposite ends: the fibre link is limited by two budgets the engineer
controls, while the telescope is limited by an atmosphere the engineer cannot
control and must instead design around.

\begin{principle}[Match the detector to the resolution the system delivers]
An imaging system delivers a resolution set by whichever limit is coarsest:
diffraction, aberration, or, for a ground telescope, atmospheric seeing. The
detector's sampling must be matched to that delivered resolution, about two
pixels across the resolution element. Finer pixels waste photons on
oversampling and coarser pixels throw away resolution the optics paid for; the
plate scale is the design variable that brings the two into register.
\end{principle}

\begin{mastery}
\textbf{Why a larger aperture buys depth faster than resolution.} The two
returns on aperture scale differently, and the difference decides what a
telescope is for. Angular resolution improves only as $1/D$ through the
Rayleigh limit, and on the ground that improvement is capped by seeing until
adaptive optics intervenes, so the resolution return on aperture is both slow
and, untreated, zero past the seeing crossover. Light-gathering improves as
$D^{2}$ because the collecting area is $\pi D^{2}/4$, and in the
background-limited regime the limiting flux for a fixed signal-to-noise ratio
falls as $1/D^{2}$, which in magnitudes is $\Delta m = 2.5\log_{10}(D_{2}^{2}/
D_{1}^{2}) = 5\log_{10}(D_{2}/D_{1})$. Doubling the aperture therefore reaches
$5\log_{10}2 = 1.5$ magnitudes deeper, a factor of four in flux, every time.
The asymmetry is why the headline number for a survey or a deep-field
instrument is always the aperture and the integration time, not the
diffraction limit: photons, gathered as $D^{2}$, are what a large telescope
mostly buys, and resolution is a bonus that the sky grants only at the best
sites or through a deformable mirror. The same scaling read in reverse explains
why space telescopes, freed of seeing, justify their cost on resolution as well
as depth: above the atmosphere the $1/D$ resolution return is real, and a
$2.4\,\si{\meter}$ Hubble delivers its full diffraction limit where a
$2.4\,\si{\meter}$ ground telescope would not.
\end{mastery}

\section{A third end-to-end case study: a time-of-flight LIDAR}
\pathtag{standard}

The fibre link carried the wave-optics budget and the telescope carried the
imaging budget; a scanning LIDAR carries the active-sensing budget, where the
engineer supplies the light, watches a fraction of it return, and converts a
photon count into a range. We design a short-range scanning time-of-flight
LIDAR for an autonomous vehicle from a sensing requirement, working through
the pulse energy and shape, the beam divergence, the range equation that ties
returned power to target reflectance and distance, the detector
signal-to-noise ratio, the maximum range and the range resolution, the
eye-safety limit that constrains the whole design, and a verdict an instrument
engineer would sign. Every laser and detector move of the chapter appears once:
Gaussian-beam divergence, the inverse-square fall of collected power,
shot-noise-limited detection, and the timing precision that fixes resolution.

\subsection{The requirement and the source}

A vehicle sensor must range opaque road targets out to $100\,\si{\meter}$ with
a depth resolution of $5\,\si{\centi\meter}$ and an angular sampling of
$0.1^{\circ}$, scanning a $30^{\circ}$ field at $10\,\si{\hertz}$, and it must
be eye-safe for bystanders. The requirement sets three quantities at once: the
pulse energy, which must put enough photons on the most distant, darkest target
to clear the detector noise; the pulse width, which sets the depth resolution;
and the beam divergence, which must stay tight enough that the spot on a
$100\,\si{\meter}$ target is smaller than the angular sample.

A near-infrared pulsed laser diode at $\lambda = 905\,\si{\nano\meter}$ is the
working choice, where silicon detectors are sensitive and the wavelength sits
in an eye-safety regime the verdict will test. The transmitter emits
$E_{p} = 1\,\si{\micro\joule}$ in a pulse of width
$t_{p} = 3\,\si{\nano\second}$, so the peak power is
$P_{\text{pk}} = E_{p}/t_{p} = 10^{-6}/(3\times10^{-9}) = 333\,\si{\watt}$
during the pulse, while the average power at a pulse-repetition rate that fills
the scan stays modest. The scan needs
$(30^{\circ}/0.1^{\circ})\times 10\,\si{\hertz} = 3000\,\text{samples}\times
10 = 3\times10^{4}$ pulses per second along one line, and a realistic raster
runs near $2\times10^{5}\,\si{\hertz}$ pulse-repetition rate, giving an average
power $E_{p}\times f_{\text{rep}} = 10^{-6}\times 2\times10^{5} =
0.2\,\si{\watt}$. Figure \ref{fig:vol04ch11:lidar-pulse-timing} shows the
transmit pulse and its delayed, attenuated return.

\input{volumes/04-energy/ch11-optics/figures/fig-lidar-pulse-timing.tex}

\subsection{Beam divergence and the spot on target}

The transmitter beam diverges by the Gaussian limit the worked examples set,
$\theta \approx \lambda/(\pi w_{0})$, but a practical LIDAR collimates with a
small lens to a fixed full-angle divergence near
$\theta_{\text{full}} = 3\,\si{\milli\radian}$, set by the trade between a
tighter beam (better angular resolution, harder alignment) and a looser one
(eye-safety relief, easier scanning). At $R = 100\,\si{\meter}$ the spot
diameter is
\[
d_{\text{spot}} = \theta_{\text{full}}\,R
= 3\times10^{-3}\times 100 = 0.30\,\si{\meter}.
\]
The angular sample of $0.1^{\circ} = 1.7\,\si{\milli\radian}$ subtends
$0.17\,\si{\meter}$ at $100\,\si{\meter}$, so the $0.30\,\si{\meter}$ spot
slightly overfills the sample and blurs the angular resolution at the far edge;
tightening the divergence to $1.5\,\si{\milli\radian}$ would match the spot to
the sample, a refinement the verdict notes. The spot smaller than most road
targets means the beam overfills small objects and underfills large flat ones,
which sets which term of the range equation governs the return.

\subsection{The range equation and the returned power}

The fraction of transmitted power that returns is the heart of an active
sensor's budget. For a target larger than the beam (a road surface, a vehicle
flank) that scatters as a Lambertian (diffuse) reflector of reflectance
$\rho$, the returned power collected by a receiver aperture of diameter
$D_{r}$ at range $R$ is
\[
P_{r} = P_{t}\,\rho\,\frac{A_{r}}{\pi R^{2}}\,\eta_{\text{atm}}\,
\eta_{\text{sys}},
\qquad A_{r} = \frac{\pi D_{r}^{2}}{4},
\]
where the $1/(\pi R^{2})$ factor is the Lambertian scatter into a hemisphere
seen through the receiver solid angle $A_{r}/R^{2}$, $\eta_{\text{atm}}$ is the
two-way atmospheric transmission, and $\eta_{\text{sys}}$ the optical
throughput. Take the worst case the requirement implies: a dark target of
reflectance $\rho = 0.1$ (asphalt, dark clothing) at $R = 100\,\si{\meter}$, a
receiver of $D_{r} = 50\,\si{\milli\meter}$
($A_{r} = \pi(0.025)^{2} = 1.96\times10^{-3}\,\si{\meter\squared}$), clear-air
two-way transmission $\eta_{\text{atm}} = 0.9$, and optical throughput
$\eta_{\text{sys}} = 0.6$. With peak transmit power
$P_{t} = 333\,\si{\watt}$,
\[
P_{r} = 333\times 0.1\times
\frac{1.96\times10^{-3}}{\pi\,(100)^{2}}\times 0.9\times 0.6
= 333\times 0.1\times 6.24\times10^{-8}\times 0.54
= 1.1\times10^{-6}\,\si{\watt}.
\]
About a microwatt returns during the pulse, eight orders of magnitude below
the transmitted peak, the steep cost of the round trip that every active
sensor pays.

\subsection{Detector signal-to-noise and the maximum range}

A microwatt of $905\,\si{\nano\meter}$ light carries a photon rate
$P_{r}/(hc/\lambda) = 1.1\times10^{-6}/(2.2\times10^{-19}) =
5.0\times10^{12}\,\text{photons}/\si{\second}$, and over the
$3\,\si{\nano\second}$ pulse that is
$1.5\times10^{4}$ photons. A silicon avalanche photodiode of quantum
efficiency $0.4$ converts these to $N = 6\times10^{3}$ primary photoelectrons.
In the shot-noise-limited regime the signal-to-noise ratio is
\[
\text{SNR} = \frac{N}{\sqrt{N + N_{b} + (\sigma_{r}/M)^{2}}}
\approx \sqrt{N} = \sqrt{6\times10^{3}} = 77,
\]
when the daylight-background count $N_{b}$ and the avalanche-multiplied read
noise stay below the signal, which a narrow optical bandpass filter and a short
range gate ensure. An SNR of $77$ at $100\,\si{\meter}$ against the
$\rho = 0.1$ target clears the detection threshold of $\text{SNR}\approx 6$
with a wide margin. The return SNR falls as $1/R^{2}$ for this overfilled-spot
geometry, so the range at which it would drop to the threshold is
$R_{\max} = 100\sqrt{77/6} = 100\times 3.6 = 360\,\si{\meter}$ against the same
dark target, comfortably past the $100\,\si{\meter}$ requirement, with the
headroom absorbing rain, dirt on the window, and a darker target. Figure
\ref{fig:vol04ch11:lidar-range-snr} carries the SNR down the range.

\input{volumes/04-energy/ch11-optics/figures/fig-lidar-range-snr.tex}

\subsection{Range resolution and timing}

Range follows from the round-trip delay, $R = c\tau/2$, so range resolution is
timing resolution scaled by $c/2$. A single-shot timing precision tied to the
pulse width gives
$\delta R \approx c\,t_{p}/2 = 3\times10^{8}\times 3\times10^{-9}/2 =
0.45\,\si{\meter}$, eight times coarser than the $5\,\si{\centi\meter}$
requirement. Two standard moves close the gap: a constant-fraction or
centroid timing discriminator locates the pulse centre to a fraction
$1/\text{SNR}$ of its width, sharpening the precision to
$\delta R \approx (c\,t_{p}/2)/\text{SNR} = 0.45/77 = 6\,\si{\milli\meter}$ on
a strong return; and averaging $n$ pulses per sample improves it as
$1/\sqrt{n}$. The requirement's $5\,\si{\centi\meter}$ resolution is met on the
$\rho = 0.1$ target at $100\,\si{\meter}$ by centroid timing alone, with
averaging in reserve for weaker returns. The pulse width therefore sets a
floor that the signal-to-noise ratio is spent buying back, the same trade the
telescope spent aperture on and the fibre link spent dispersion compensation
on.

\subsection{Eye safety: the constraint that sizes the laser}

A LIDAR points an infrared laser at public space, so the maximum permissible
exposure governs the pulse energy as hard as the range equation does. At
$905\,\si{\nano\meter}$ the eye still focuses the beam onto the retina, but the
near-infrared is absorbed more by the ocular media than visible light, which
relaxes the limit relative to $1\,\si{\micro\meter}$. The exposure that matters
is the energy a $7\,\si{\milli\meter}$ pupil collects from a single pulse at
the closest a bystander can approach the aperture. A pupil at the exit window
gathers
$E_{\text{eye}} = E_{p}\,(A_{\text{pupil}}/A_{\text{beam}})$; with the beam
expanded over the transmitter aperture the collected fraction is well below
unity, and the per-pulse retinal exposure stays under the
$905\,\si{\nano\meter}$ single-pulse limit. The binding constraint is the
\emph{repetitive}-pulse exposure: a scanning beam can dwell on a stationary eye
for the scan revisit time, so the limit drops by a factor $n^{-1/4}$ in the
number of pulses delivered during the eye's $0.25\,\si{\second}$ aversion
response. The design holds eye-safe by keeping the per-pulse energy at
$1\,\si{\micro\joule}$, expanding the beam over a $25\,\si{\milli\meter}$ exit
aperture so the on-axis irradiance at the window is low, and scanning fast
enough that no eye sees a damaging pulse train, the combination that puts the
device in the eye-safe Class 1 envelope under normal operation. A higher pulse
energy would buy range but breach the limit; eye safety, not the photon budget,
caps the transmitter.

\subsection{Verdict}

The sensor closes against its requirement. A $1\,\si{\micro\joule}$,
$3\,\si{\nano\second}$, $905\,\si{\nano\meter}$ pulse through a
$3\,\si{\milli\radian}$ beam returns about a microwatt from a $\rho = 0.1$
target at $100\,\si{\meter}$, which a quantum-efficiency-$0.4$ avalanche
photodiode detects at $\text{SNR}\approx 77$, clearing the threshold to a
maximum range near $360\,\si{\meter}$ on the same dark target. Centroid timing
sharpens the $0.45\,\si{\meter}$ pulse-width floor to a few centimetres,
meeting the $5\,\si{\centi\meter}$ depth requirement. Eye safety, not range,
sets the pulse energy, and the expanded exit aperture with fast scanning keeps
the device Class 1. The one refinement the budget flags is the beam
divergence: at $3\,\si{\milli\radian}$ the far-field spot slightly overfills
the angular sample, so a production design tightens the divergence to
$1.5\,\si{\milli\radian}$ to recover angular resolution at the cost of tighter
alignment. The instrument that ships is a $905\,\si{\nano\meter}$ diode
transmitter, a $50\,\si{\milli\meter}$ receiver with a narrow bandpass filter
and range gate, an avalanche-photodiode detector with constant-fraction
timing, and a fast scanner, with the range, resolution, and eye-safety budgets
all cleared and documented. The three case studies make one combined point:
the fibre link is capped by two budgets the engineer controls, the telescope
by an atmosphere the engineer cannot control, and the LIDAR by a safety limit
the engineer must not exceed, so each design closes against a different kind of
ceiling.

\begin{principle}[An active sensor pays the round trip twice]
An active optical sensor supplies its own illumination and collects only the
small fraction that scatters back, so its returned power falls as the product
of a one-way transmit term and a one-way collection term, an inverse-square
law in the overfilled-spot regime and a steeper $1/R^{4}$ when the beam
underfills a small target. The pulse width sets the range resolution before
timing discrimination buys it back, and for any sensor pointed at public space
the eye-safety limit, not the photon budget, sizes the source.
\end{principle}

\begin{mastery}
\textbf{Point targets, the $1/R^{4}$ law, and the radar analogy.} The
range equation splits into two regimes that scale differently, and confusing
them mis-sizes a sensor by orders of magnitude. When the beam \emph{overfills}
the target, a fixed-size object intercepts a fraction of the transmitted power
that itself falls as $1/R^{2}$ (the spot grows as $R^{2}$), and the scattered
light is then collected through a receiver solid angle that falls as a second
$1/R^{2}$, so the returned power obeys an overall $1/R^{4}$ law, the
point-target (radar) regime. When the beam \emph{underfills} an extended
target, all the transmitted power lands on the target and only the receiver
solid-angle term survives, giving the $1/R^{2}$ law the case study used for the
road surface. The crossover is the range at which the spot equals the target,
$R_{\text{cross}}\approx d_{\text{target}}/\theta_{\text{full}}$; a
$0.3\,\si{\meter}$ target at $\theta_{\text{full}}=3\,\si{\milli\radian}$
crosses at $100\,\si{\meter}$, exactly the requirement boundary, which is why
the design sits on the seam between the two laws and a thin wire at
$100\,\si{\meter}$ returns far less than the road does. The same $1/R^{4}$
bookkeeping governs microwave radar, where it is the textbook form; optical
LIDAR inherits it and adds the diffraction-limited beam that makes the
overfilled-extended-target regime reachable at all, the advantage a pencil
beam buys over a radar's broad lobe.
\end{mastery}

\section{Failure: aberrations that ruined a mission}\pathtag{core}

The Hubble Space Telescope launched on 24 April 1990 aboard Space
Shuttle Discovery (STS-31). Within weeks of the first imaging tests,
NASA confirmed that the primary mirror had been figured to the wrong
shape: the outer edge was approximately $2\,\si{\micro\meter}$ too
flat for the design parabola, an error large enough to defeat
diffraction-limited imaging across the telescope's bandpass
\cite{acc:nasa-hubble-optical-systems-1990}.

The NASA Hubble Space Telescope Optical Systems Failure Report,
prepared by the Optical Systems Board of Investigation under Lew
Allen Jr.\ and released in November 1990, traced the error to a
misassembly of the reflective null corrector (RNC), the optical test
instrument used by Perkin-Elmer Corporation to measure the primary
mirror's figure during polishing. A field lens within the RNC was
mispositioned along the optical axis by approximately $1.3\,\si{\milli
\meter}$, the result of a misinterpretation of a measurement target
on a metering rod during the RNC's assembly: the end cap of the rod
and the target feature were not properly distinguished, and the
spacing was set against the wrong feature. The error in the test
instrument propagated into the mirror's polished figure with the
authority of an authoritative test result. Two auxiliary optical
tests, a refractive null corrector and an inverse null test, had
indicated discrepancies during fabrication; both were set aside on
the assumption that the RNC was more authoritative
\cite{acc:nasa-hubble-optical-systems-1990}.

The on-orbit aberration was spherical aberration of $\sim \lambda/2$
peak-to-valley at $550\,\si{\nano\meter}$, severe enough to spread
the Airy disc into a halo of diameter $\sim 1$ arc-second and to
reduce the Strehl ratio (the on-axis peak intensity divided by the
diffraction-limited peak) from a target of $\sim 0.85$ to
approximately $0.15$. Optical correction was delivered by COSTAR
(Corrective Optics Space Telescope Axial Replacement), installed
during the STS-61 servicing mission in December 1993. COSTAR was a
set of small mirrors placed in the optical paths of the
non-camera instruments, figured to cancel the primary's
aberration. The Wide Field/Planetary Camera 2, installed at the same
servicing mission, had its corrective optics integrated into the
instrument itself, the modern approach.

The failure was not the optician's; the Perkin-Elmer team polished
the mirror to the figure their measurement instrument said the
design required. The failure was the absence of process barriers
that would have caused the measurement instrument itself to be
verified, that would have required the auxiliary tests' anomalies to
be formally resolved before fabrication proceeded, and that would
have given NASA's quality-assurance representation effective
oversight of the test apparatus rather than only of the test results.
The Allen Commission's recommendations centred on these missing
barriers: independent verification of test instruments, formal
anomaly-resolution procedures that cannot be informally overridden,
and engaged customer oversight of the test apparatus
\cite{acc:nasa-hubble-optical-systems-1990}.

The Hubble case is the canonical metrology-driven failure in
twentieth-century optical engineering. The lesson by scale of this
chapter is twofold. First, an instrument whose measurement chain is
not independently verified is not a reference; it is a measurement.
Second, when two tests disagree, ``the more authoritative one'' is a
process answer rather than a physical answer; the engineer must
resolve the disagreement before fabricating to either reading.

\begin{principle}[An unverified test instrument is a measurement]
A test instrument that is itself the only authority on its own
measurement chain is not a reference; it is a measurement subject to
the same uncertainty as the artifact it measures. The discipline of
the optical-fabrication shop, and the metrology-traceability
discipline of Volume I, requires that every test instrument's
measurement chain be independently verified before the instrument is
trusted to certify another artifact's figure.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Build a Michelson interferometer and measure a small
displacement]
Build a Michelson interferometer on an optical bench and use it to
measure a small displacement at the wavelength precision the
instrument is capable of.

\textbf{Materials.}
A laboratory laser (a $5\,\si{\milli\watt}$, Class~2 or 3R helium-neon
or stabilised diode laser at $632.8\,\si{\nano\meter}$ or
$635\,\si{\nano\meter}$); a $1$-inch beamsplitter cube; two
front-surface mirrors mounted on kinematic bases; a piezoelectric
actuator with a calibrated displacement range of $\sim 10\,\si{
\micro\meter}$ (a stack actuator with $\sim 1\,\si{\volt\per\micro
\meter}$ sensitivity is appropriate); a CCD or smartphone camera as
the imaging detector; an optical breadboard.

\textbf{Procedure.}
\begin{enumerate}
  \item Align the laser, beamsplitter, and two mirrors so that the
    two reflected beams overlap on the detector with a tilt that
    produces fringes of $\sim 1\,\si{\milli\meter}$ spacing.
  \item Adjust the path lengths to within a few millimetres of equal,
    minimising the effect of laser coherence length.
  \item Mount the piezoelectric actuator on the back of one mirror's
    kinematic base so that voltage on the piezo translates the
    mirror along the optical axis.
  \item Apply a triangular voltage waveform to the piezo from $0$ to
    $\pm 10\,\si{\volt}$ and record the fringe motion on the
    detector.
  \item Count the number of fringes that pass a fixed reference line
    per volt of piezo drive. Each fringe corresponds to $\lambda/2$
    of mirror displacement ($316\,\si{\nano\meter}$ for a HeNe).
\end{enumerate}

\textbf{Report.} A one-page report with the setup photograph, the
fringe-counting data, the inferred piezo sensitivity in $\si{\nano
\meter\per\volt}$, the manufacturer's specification, and a discussion
of the discrepancy. Expected: the measured piezo sensitivity should
match the manufacturer's specification to within $\sim 10\,\%$. The
dominant error sources are interferometer-arm mechanical vibration
(reduce by mounting on a vibration-isolated breadboard), thermal
drift of the optical paths (reduce by short measurement intervals),
and detector noise (reduce by averaging multiple sweeps).
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A thin biconvex lens has surface radii $R_{1} = +75\,\si{\milli\meter}$,
$R_{2} = -50\,\si{\milli\meter}$ in BK7 glass ($n = 1.517$ at
$\lambda = 587\,\si{\nano\meter}$). Compute the focal length.
\paragraph{Solution sketch.} Lens-maker's equation
$1/f = (n-1)(1/R_{1} - 1/R_{2}) = 0.517\,(1/75 - 1/(-50)) = 0.517\,
(0.01333 + 0.02000) = 0.517\times 0.03333 = 0.01723\,\si{\per\milli
\meter}$, so $f = 58.0\,\si{\milli\meter}$. The two convex surfaces add
their curvatures because $R_{2}$ is negative. Matches
\texttt{code/thin\_lens.py}.
\end{exercise}

\begin{exercise}
An object is placed $30\,\si{\centi\meter}$ in front of a converging
thin lens of focal length $20\,\si{\centi\meter}$. Compute the image
distance, the lateral magnification, and whether the image is real
or virtual, erect or inverted.
\paragraph{Solution sketch.} Thin-lens equation
$1/s_{i} = 1/f - 1/s_{o} = 1/20 - 1/30 = 1/60$, so
$s_{i} = 60\,\si{\centi\meter}$. Magnification
$M = -s_{i}/s_{o} = -60/30 = -2$. The positive $s_{i}$ marks a real
image; the negative $M$ marks inversion; $|M| = 2$ marks twofold
enlargement. This is the case drawn in Figure
\ref{fig:vol04ch11:lens-ray-diagram}.
\end{exercise}

\begin{exercise}
A $200\,\si{\milli\meter}$-diameter telescope objective operates at
$550\,\si{\nano\meter}$. Compute the angular resolution (Rayleigh
criterion) in arc-seconds and the minimum separation it can resolve
on the lunar surface ($384\,000\,\si{\kilo\meter}$ distant).
\paragraph{Solution sketch.} Rayleigh angle
$\theta = 1.22\lambda/D = 1.22\times 550\times10^{-9}/0.200 = 3.36\times
10^{-6}\,\si{\radian}$. Converting, $3.36\times10^{-6}\times
(180/\pi)\times 3600 = 0.69$ arc-second. The resolved separation on the
Moon is $\theta R = 3.36\times10^{-6}\times 3.84\times10^{8} = 1.3\,
\si{\kilo\meter}$. A backyard $200\,\si{\milli\meter}$ telescope, if
seeing allowed, could just separate two features $1.3\,\si{\kilo\meter}$
apart on the Moon. Matches \texttt{code/diffraction.py}.
\end{exercise}

\begin{exercise}
A Young's-double-slit experiment uses two slits separated by
$0.5\,\si{\milli\meter}$ at $\lambda = 633\,\si{\nano\meter}$ on a
screen $2\,\si{\meter}$ away. Compute the fringe spacing.
\paragraph{Solution sketch.} Fringe spacing
$\Delta y = \lambda L/d = (633\times10^{-9})(2)/(0.5\times10^{-3}) =
2.53\times10^{-3}\,\si{\meter} = 2.5\,\si{\milli\meter}$. The fringes
are millimetre-scale and easily seen, which is why the double slit is
the classroom demonstration of the wave nature of light. Halving the
slit separation would double the spacing.
\end{exercise}

\begin{exercise}
A microscope objective is rated NA = $0.65$ at the standard $4.7\,\si{
\milli\meter}$ working distance in air. Compute the resolution at
$\lambda = 500\,\si{\nano\meter}$ and the half-angle of the
illumination cone.
\paragraph{Solution sketch.} Resolution
$d_{\min} = 0.61\lambda/\text{NA} = 0.61\times 500\times10^{-9}/0.65 =
4.7\times10^{-7}\,\si{\meter} = 470\,\si{\nano\meter}$. Half-angle from
$\text{NA} = n\sin\theta_{\max}$ in air ($n = 1$):
$\theta_{\max} = \arcsin(0.65) = 40.5^{\circ}$. A dry objective cannot
exceed $\text{NA} = 1$; reaching $470\,\si{\nano\meter}$ resolution
needs no immersion, but finer work does.
\end{exercise}

\begin{exercise}
A single-mode optical fibre has $n_{\text{core}} = 1.4500$,
$n_{\text{clad}} = 1.4467$. Compute the numerical aperture, the
critical angle (from inside the core), and the maximum acceptance
angle in air at the fibre input face.
\paragraph{Solution sketch.}
$\text{NA} = \sqrt{n_{\text{core}}^{2} - n_{\text{clad}}^{2}} =
\sqrt{1.4500^{2} - 1.4467^{2}} = \sqrt{2.1025 - 2.0929} =
\sqrt{0.00957} = 0.0978$. Critical angle (inside core)
$\theta_{c} = \arcsin(n_{\text{clad}}/n_{\text{core}}) =
\arcsin(0.99772) = 86.1^{\circ}$, so rays must strike the boundary
within $3.9^{\circ}$ of grazing to be guided. Acceptance half-angle in
air $\theta_{a} = \arcsin(\text{NA}) = \arcsin(0.0978) = 5.6^{\circ}$.
Matches \texttt{code/fiber\_na.py}.
\end{exercise}

\begin{exercise}
A laser of wavelength $785\,\si{\nano\meter}$ has a linewidth of
$10\,\si{\mega\hertz}$. Compute the coherence length and the
maximum interferometer-arm imbalance for which fringes can still
form.
\paragraph{Solution sketch.} Coherence length
$L_{c} \approx c/\Delta f = (3\times10^{8})/(10\times10^{6}) =
30\,\si{\meter}$. Fringe contrast falls off as the path imbalance
approaches $L_{c}$; usable fringes persist while the round-trip
imbalance is well within $L_{c}$, so the arm-length difference (half the
round trip) can be up to about $15\,\si{\meter}$. The linewidth, not the
wavelength, sets how far apart the arms may be, which is why narrow-line
lasers are the metrology choice.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the thin-lens equation $1/s_{o} + 1/s_{i} = 1/f$ by applying
the small-angle refraction equation $n_{1}/s_{o} + n_{2}/s_{i} =
(n_{2}-n_{1})/R$ at both surfaces of a thin lens in air. State
precisely where the thin-lens approximation is used.
\paragraph{Solution sketch.} Apply the single-surface refraction
relation at surface~1 (air to glass, radius $R_{1}$):
$1/s_{o} + n/s_{i1} = (n-1)/R_{1}$, where $s_{i1}$ is the intermediate
image distance. The intermediate image serves as the object for
surface~2 (glass to air, radius $R_{2}$); for a lens of zero thickness
the object distance there is $-s_{i1}$ (a virtual object on the far
side), giving $-n/s_{i1} + 1/s_{i} = (1-n)/R_{2}$. Adding the two
equations cancels the $s_{i1}$ terms and yields
$1/s_{o} + 1/s_{i} = (n-1)(1/R_{1} - 1/R_{2}) = 1/f$. The thin-lens
approximation enters at exactly one step: setting the object distance
for surface~2 equal to $-s_{i1}$, which assumes the lens thickness is
negligible compared with the image distances.
\end{exercise}

\begin{exercise}
Derive the Rayleigh resolution criterion $\theta_{\min} = 1.22\lambda
/D$ for a circular aperture by treating the aperture function as
unity inside a disc and zero outside, computing its Fraunhofer
diffraction pattern, and identifying the first zero of the resulting
Bessel function.
\paragraph{Solution sketch.} The Fraunhofer pattern is the
two-dimensional Fourier transform of the aperture function. For a
uniform disc of radius $a$, the transform is the circularly symmetric
integral $\int_{0}^{a} J_{0}(kr\sin\theta)\,r\,dr$, which evaluates to
$\propto J_{1}(ka\sin\theta)/(ka\sin\theta)$ using the identity
$\int_{0}^{x} J_{0}(u)u\,du = x J_{1}(x)$. The intensity is the square
of this amplitude, the Airy pattern. The first zero is the first zero of
$J_{1}$, at argument $3.832$. Setting
$ka\sin\theta = (2\pi/\lambda)(D/2)\sin\theta = 3.832$ and solving gives
$\sin\theta = 3.832\lambda/(\pi D) = 1.22\lambda/D$, since
$3.832/\pi = 1.22$. The Rayleigh criterion places the second source's
peak at this first zero.
\end{exercise}

\begin{exercise}
Derive the lens-maker's equation $1/f = (n-1)(1/R_{1} - 1/R_{2})$
from the small-angle refraction relation, applied twice.
\paragraph{Solution sketch.} Send parallel light ($s_{o} \to \infty$)
into the thin lens; the image forms at the focal point, $s_{i} = f$. The
two-surface derivation of the previous problem reduces, with
$1/s_{o} = 0$, to $1/f = (n-1)(1/R_{1} - 1/R_{2})$ directly. The same
cancellation of the intermediate image distance does the work; the
infinite-object case isolates $f$ without carrying the imaging terms,
and the result holds for finite objects because $f$ is a property of the
lens alone.
\end{exercise}

\begin{exercise}
Derive the fringe-spacing formula $\Delta y = \lambda L/d$ for
Young's-double-slit setup, starting from the path-length difference
between the two slits and a far-field point on the screen.
\paragraph{Solution sketch.} The path difference from the two slits to a
screen point at angle $\theta$ is $\delta = d\sin\theta$. Bright fringes
require $\delta = m\lambda$, so $\sin\theta_{m} = m\lambda/d$. For
$L \gg d$ the small-angle approximation gives
$\sin\theta \approx \tan\theta = y/L$, so the $m$th bright fringe sits
at $y_{m} = m\lambda L/d$. Adjacent fringes differ by $m \to m+1$, so
$\Delta y = y_{m+1} - y_{m} = \lambda L/d$. The spacing is uniform
because the path difference is linear in $y$ in this approximation. See
Figure \ref{fig:vol04ch11:young-double-slit}.
\end{exercise}

\begin{exercise}
Derive the numerical-aperture formula $\text{NA} = \sqrt{n_{\text{
core}}^{2}-n_{\text{clad}}^{2}}$ for a step-index optical fibre, by
considering a ray at the critical angle for total internal reflection
at the core-cladding boundary and tracing it back to the fibre's
input face.
\paragraph{Solution sketch.} A ray guided by total internal reflection
strikes the core-cladding boundary at the critical angle
$\theta_{c}$, where $\sin\theta_{c} = n_{\text{clad}}/n_{\text{core}}$.
The complementary angle to the fibre axis is
$\phi = 90^{\circ} - \theta_{c}$, with
$\sin\phi = \cos\theta_{c} = \sqrt{1 - (n_{\text{clad}}/n_{\text{core}})
^{2}}$. Refraction at the input face (air to core) gives
$\sin\theta_{a} = n_{\text{core}}\sin\phi$ by Snell's law, so
$\sin\theta_{a} = n_{\text{core}}\sqrt{1 - n_{\text{clad}}^{2}/
n_{\text{core}}^{2}} = \sqrt{n_{\text{core}}^{2} - n_{\text{clad}}^{2}}$.
That is the numerical aperture, $\text{NA} = \sin\theta_{a}$. See Figure
\ref{fig:vol04ch11:fiber-tir}.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the minimum reliably-detected angular separation of two
stars by the unaided human eye (pupil diameter $\sim 4\,\si{\milli
\meter}$, peak visual wavelength $550\,\si{\nano\meter}$). Compare
the diffraction limit to the empirical visual acuity of about
$1$ arc-minute.
\paragraph{Solution sketch.} Diffraction limit
$\theta = 1.22\lambda/D = 1.22\times 550\times10^{-9}/0.004 =
1.7\times10^{-4}\,\si{\radian}$, which is
$1.7\times10^{-4}\times(180/\pi)\times 60 = 0.58$ arc-minute. The eye's
empirical acuity of about $1$ arc-minute is within a factor of two of
the diffraction limit, so the eye is close to diffraction-limited at the
centre of the field. Retinal cone spacing (about $2\,\si{\micro\meter}$
at the fovea, subtending roughly $0.5$ arc-minute) is matched to the
diffraction limit, the expected outcome of a well-engineered detector.
\end{exercise}

\begin{exercise}
Estimate the temporal coherence length of sunlight falling on the
ground, given that visible sunlight covers a wavelength range from
$\sim 400$ to $\sim 700\,\si{\nano\meter}$. State the implication for
laboratory interferometry without a narrowband filter.
\paragraph{Solution sketch.} The frequency spread of visible sunlight is
$\Delta f = c(1/\lambda_{\min} - 1/\lambda_{\max}) = 3\times10^{8}\,
(1/400\times10^{-9} - 1/700\times10^{-9}) \approx 3.2\times10^{14}\,\si{
\hertz}$. The coherence length is
$L_{c} \approx c/\Delta f \approx 3\times10^{8}/3.2\times10^{14} \approx
0.9\,\si{\micro\meter}$, of order one wavelength. Interferometry with
white light works only when the two arms are matched to within about a
micron, which is why white-light fringes are used to locate the
zero-path-difference point precisely, and why a narrowband filter is
needed to see fringes at any appreciable path imbalance.
\end{exercise}

\begin{exercise}
Estimate the depth of field of a $50\,\si{\milli\meter}$ camera lens
focused at $5\,\si{\meter}$ at $f/4$, with the circle of confusion
set to $30\,\si{\micro\meter}$ (the diagonal divided by $\sim 1500$
for a full-frame sensor).
\paragraph{Solution sketch.} The hyperfocal distance is
$H \approx f^{2}/(Nc) = (0.050)^{2}/(4\times 30\times10^{-6}) =
0.0025/1.2\times10^{-4} = 20.8\,\si{\meter}$. With the lens focused at
$s = 5\,\si{\meter}$, the near and far limits are
$D_{n} = sH/(H+s) = 5\times 20.8/25.8 = 4.0\,\si{\meter}$ and
$D_{f} = sH/(H-s) = 5\times 20.8/15.8 = 6.6\,\si{\meter}$, a total depth
of field of about $2.6\,\si{\meter}$. The depth is asymmetric, extending
farther behind the focus than in front, the familiar one-third-in-front
rule of thumb.
\end{exercise}

\begin{exercise}
Estimate the photon-shot-noise limit of a $14$-bit, $5\,\si{\micro
\meter}$-pixel scientific CMOS sensor with $50\,000$-electron full
well at maximum brightness. State the consequent dynamic-range
limit and the read-noise floor that would be acceptable.
\paragraph{Solution sketch.} At full well $N = 50\,000$ electrons, the
shot noise is $\sqrt{N} = 224$ electrons, so the signal-to-noise ratio
at saturation is $50\,000/224 = 224$. The shot-noise-limited dynamic
range is therefore about $224{:}1$, or $47\,\si{\decibel}$. Mapping the
full well onto $14$ bits gives $50\,000/16384 = 3.05$ electrons per
digital level, so a read-noise floor near $3$ electrons matches the
quantisation and the digitiser is well chosen; a read noise much below
$3$ electrons would be wasted because quantisation dominates, and much
above would throw away the sensor's low-light reach.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a Fresnel-propagation algorithm using the
angular-spectrum method to propagate a Gaussian beam from a $\lambda
/100$ initial spot to a $1\,\si{\meter}$-distant observation plane.
Verify against the closed-form Gaussian-beam propagation and
identify the regime in which the Fresnel approximation begins to
break down.
\paragraph{Solution sketch.} The angular-spectrum method takes the
two-dimensional Fourier transform of the initial field, multiplies by
the free-space transfer function $H(f_{x},f_{y}) = \exp(i k z
\sqrt{1 - (\lambda f_{x})^{2} - (\lambda f_{y})^{2}})$, and
inverse-transforms. Sample the input plane finely enough that the
highest spatial frequency $\text{NA}/\lambda$ is below the Nyquist
limit; for a $\lambda/100$ spot this demands a grid pitch well under
$\lambda$. Compare the propagated waist and on-axis amplitude with the
closed-form Gaussian beam, $w(z) = w_{0}\sqrt{1 + (z/z_{R})^{2}}$ with
$z_{R} = \pi w_{0}^{2}/\lambda$. The Fresnel (paraxial) approximation
expands the square root to first order and breaks down when
$(\lambda f)^{2}$ is no longer small, that is for high-NA beams or very
short propagation where evanescent and wide-angle components matter;
agreement should be excellent for the modest-NA Gaussian and degrade as
$w_{0}$ shrinks toward $\lambda$.
\end{exercise}

\begin{exercise}
Implement a ray-tracer for a thin lens with spherical aberration,
sampling rays uniformly across the aperture. Plot the spot diagram
at the paraxial focus, the marginal focus, and the circle of least
confusion. Identify the empirical optimum focus position relative to
the paraxial focus.
\paragraph{Solution sketch.} For a thin lens with spherical aberration,
a ray entering at radial height $r$ crosses the axis at a longitudinal
position shifted from the paraxial focus by $\Delta z \propto r^{2}$
(transverse aberration $\propto r^{3}$). Sample rays uniformly in $r$,
trace each to the chosen image plane, and plot the intersection points.
At the paraxial focus the marginal rays form a large outer ring around a
bright core; at the marginal focus the core blurs and the marginal rays
tighten. The circle of least confusion, the smallest enclosing spot,
sits about three-quarters of the way from the paraxial focus toward the
marginal focus for third-order spherical aberration; that intermediate
plane is the empirical best focus, and its spot radius is roughly a
quarter of the paraxial-plane ring radius.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a $20\times$ achromatic microscope objective at a working
distance of $5\,\si{\milli\meter}$ in air. Specify the focal length,
the aperture, the expected NA, and the back-focal-plane diameter
required for an eyepiece of $f = 10\,\si{\milli\meter}$.
\paragraph{Solution sketch.} For a standard $160\,\si{\milli\meter}$
tube length, a $20\times$ objective has focal length
$f_{\text{obj}} = 160/20 = 8\,\si{\milli\meter}$. A $20\times$ objective
typically carries $\text{NA} \approx 0.40$, so the half-angle is
$\arcsin(0.40) = 23.6^{\circ}$ and the front aperture diameter is about
$2 f_{\text{obj}}\tan(23.6^{\circ}) \approx 7\,\si{\milli\meter}$. The
achromat is a cemented crown-flint doublet correcting axial colour at
the F and C lines. The back-focal-plane (exit-pupil) diameter that fills
a $10\,\si{\milli\meter}$ eyepiece is set by matching exit pupil to
eyepiece field stop, of order $4$ to $5\,\si{\milli\meter}$ for
comfortable eye relief. The working distance of $5\,\si{\milli\meter}$
is short but consistent with an $8\,\si{\milli\meter}$ focal length.
\end{exercise}

\begin{exercise}
Design an optical system to image a $1\,\si{\centi\meter}$-square
object onto a $36\times 24\,\si{\milli\meter}$ camera sensor at unit
magnification. Specify the lens focal length, the working distance,
the aperture, and the expected diffraction limit at $f/8$.
\paragraph{Solution sketch.} Unit magnification ($M = -1$) requires
object and image each at $2f$, so the object and sensor are $4f$ apart.
A $1\,\si{\centi\meter}$ object onto a $36\times24\,\si{\milli\meter}$
sensor at $M = -1$ over-fills the long axis only slightly; choose a
convenient focal length, say $f = 100\,\si{\milli\meter}$, giving object
distance $s_{o} = 200\,\si{\milli\meter}$, image distance
$s_{i} = 200\,\si{\milli\meter}$, and a working distance near
$200\,\si{\milli\meter}$. At $f/8$ the effective aperture at $M = -1$ is
$N_{\text{eff}} = N(1+|M|) = 16$, so the Airy radius is
$1.22\lambda N_{\text{eff}} = 1.22\times 550\times10^{-9}\times 16 =
10.7\,\si{\micro\meter}$. The bellows factor doubling the effective
$f$-number is the detail a unit-magnification macro design must not
omit.
\end{exercise}

\begin{exercise}
Design a single-mode-fibre link for $10\,\si{\giga\bit\per\second}$
data transmission over $80\,\si{\kilo\meter}$ at $1.55\,\si{\micro
\meter}$, using one EDFA amplifier mid-span. Specify the launch
power, the receiver sensitivity, the link budget, the amplifier gain,
and the dispersion compensation required.
\paragraph{Solution sketch.} Span loss at
$0.20\,\si{\decibel\per\kilo\meter}$ over $80\,\si{\kilo\meter}$ is
$16\,\si{\decibel}$; add $\sim 2\,\si{\decibel}$ for connectors and
splices and a $3\,\si{\decibel}$ margin, total $21\,\si{\decibel}$. With
a launch power of $+3\,\text{dBm}$ and a receiver sensitivity of
$-24\,\text{dBm}$ at $10\,\si{\giga\bit\per\second}$, the budget is
$+3 - 21 = -18\,\text{dBm}$ received, $6\,\si{\decibel}$ above
sensitivity, so a mid-span EDFA with about $12\,\si{\decibel}$ gain
restores margin and supports growth. Chromatic dispersion of standard
single-mode fibre is $\sim 17\,\si{\pico\second/(\nano\meter\,\kilo
\meter)}$; over $80\,\si{\kilo\meter}$ with a $0.1\,\si{\nano\meter}$
source linewidth the spread is $\sim 136\,\si{\pico\second}$, a
significant fraction of the $100\,\si{\pico\second}$ bit period, so a
dispersion-compensating fibre module of about $-1360\,\si{\pico\second/
\nano\meter}$ is specified to flatten it.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{standard}

\begin{exercise}
A microscope's spot diagram at a single point of the field of view
shows a triangular three-lobed pattern instead of an Airy disc.
Identify the candidate aberrations (trefoil from a three-point mirror
support, primary trefoil from a manufacturing error) and the
supplementary measurement that distinguishes them.
\paragraph{Solution sketch.} A three-lobed spot is the signature of
trefoil, a $\cos 3\phi$ wavefront error. Mount-induced trefoil from a
three-point support rotates with the optic when the mount is reclocked;
a manufacturing (figure) trefoil is frozen into the glass and stays
fixed in the optic's own frame. The distinguishing measurement is to
rotate the suspect element relative to its mount and watch the lobe
orientation: if the lobes follow the mount, the support is the cause
(relieve or re-space the contacts); if they follow the glass, the figure
is at fault (re-polish or reject). An interferogram or Shack-Hartmann
wavefront map confirms the $\cos 3\phi$ term quantitatively.
\end{exercise}

\begin{exercise}
An astronomical CCD image shows two distinct ring patterns around
bright stars: an inner Airy ring and an outer halo. Identify the
candidate optical defects (defocused secondary mirror, primary-mirror
spherical aberration, dome seeing) and the test that distinguishes
them.
\paragraph{Solution sketch.} A defocused secondary throws a symmetric
ring whose diameter changes sharply with focus and which collapses to a
point at best focus; the test is to step through focus and watch the
halo shrink to nothing. Primary-mirror spherical aberration leaves a
residual halo that cannot be focused away at any setting, the Hubble
signature; the test is that no focus position removes it, and a
star-test or interferogram shows the $r^{4}$ wavefront term. Dome (local
air) seeing produces a halo that varies in time and with thermal
conditions, worst soon after dome opening; the test is to monitor the
pattern over minutes and correlate with temperature, and it improves as
the air settles. Persistence under focus and over time separates the
three.
\end{exercise}

\begin{exercise}
A laboratory measurement of an interferometer's fringe contrast
shows that the contrast drops as the path-length difference exceeds
$10\,\si{\centi\meter}$, even though the source is specified as a
laboratory HeNe laser. Identify the candidate causes (multi-mode
laser operation, vibration washing out the fringe, mechanical drift)
and the test that distinguishes them.
\paragraph{Solution sketch.} A contrast drop that sets in at a definite
path difference is the signature of finite coherence length, which an
unstabilised multi-longitudinal-mode HeNe shows at the
$\sim 20\,\si{\centi\meter}$ scale of its mode beating; the test is to
measure contrast against path difference and look for periodic revivals
at multiples of the cavity round-trip length, the fingerprint of mode
structure. Vibration washes the fringes out at all path differences and
is time-varying, not path-dependent; the test is to isolate the bench or
shorten the integration and watch the contrast recover. Mechanical drift
moves the fringes slowly and blurs only on long exposures; the test is
to shorten the exposure and see the instantaneous contrast return. The
path-difference dependence cleanly fingers coherence length as the
cause.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Summarise the technical mechanism of the Hubble Space Telescope
primary mirror failure (1990) in three sentences, using the registry
entry's working language. State which process barrier was missing
and which auxiliary tests were set aside.
\paragraph{Solution sketch.} A field lens in the reflective null
corrector used to measure the primary mirror was mispositioned by about
$1.3\,\si{\milli\meter}$, the result of setting a spacing against the
wrong feature on a metering rod. The faulty test instrument certified
the mirror as correct while it was being polished about
$2\,\si{\micro\meter}$ too flat at the edge, producing on-orbit
spherical aberration of order $\lambda/2$. The missing process barrier
was independent verification of the test instrument's own measurement
chain; two auxiliary tests, a refractive null corrector and an inverse
null test, had flagged discrepancies and were set aside as less
authoritative than the reflective null corrector
\cite{acc:nasa-hubble-optical-systems-1990}.
\end{exercise}

\begin{exercise}
A $300\,\si{\milli\meter}$-diameter astronomical telescope shows a
visible coma pattern in off-axis stars after assembly. Identify the
candidate failure modes (decentred primary, tilted secondary, focal-
plane misalignment) and the test that would distinguish them.
\paragraph{Solution sketch.} Field-uniform coma that points the same way
everywhere off-axis indicates a misaligned (decentred or tilted) optic,
since a centred system has zero coma on-axis and coma growing linearly
with field. A decentred primary and a tilted secondary both add a
constant coma across the field; they are separated by a star test on
axis, where a centred system shows a symmetric Airy pattern: residual
on-axis coma points to the secondary tilt, while a coma that reverses
sense when the secondary is reclocked points to the secondary. Focal-
plane (detector) misalignment produces defocus or astigmatism varying
across the frame, not uniform coma; the test is to check focus at
several field points. Collimation by the standard star test, adjusting
the secondary until on-axis coma vanishes, isolates and cures the
dominant term.
\end{exercise}

\begin{exercise}
A high-power laser-cutting system fails to maintain its specified
focal-spot size after two years of operation, even though all
maintenance procedures have been followed. Identify the candidate
failure modes (thermal lensing in the focusing lens, contamination
of the output window, beam-quality degradation of the laser source)
and the working diagnostic test for each.
\paragraph{Solution sketch.} Thermal lensing in the focusing optic
shifts the focal plane as the lens heats during a cut; the diagnostic is
to measure spot size against dwell time or duty cycle and look for a
drift that grows with absorbed power, then inspect the lens for
absorbing contamination. Output-window contamination scatters and
absorbs power at the window; the diagnostic is a power-meter reading
before and after the window plus a microscope inspection for deposits or
pitting. Source beam-quality degradation raises the beam parameter
product $M^{2}$, enlarging the focal spot independently of the delivery
optics; the diagnostic is a beam-profiler $M^{2}$ measurement at the
source output compared against the commissioning value. The
power-dependent (thermal), window-localised (contamination), and
source-intrinsic ($M^{2}$) signatures each point to one element.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A consulting client must decide between an aspheric singlet and a
multi-element achromat for a $5\,\si{\milli\meter}$-focal-length
collimator at $\lambda = 633\,\si{\nano\meter}$. Identify the trade-
offs in terms of cost, performance, weight, and field of view.
\paragraph{Solution sketch.} At a single design wavelength
($633\,\si{\nano\meter}$) chromatic correction is not needed, so the
aspheric singlet wins on weight, element count, and on-axis spot size,
and a moulded asphere is cheap in volume though costly in small runs.
The achromat carries no advantage at one wavelength but tolerates a
spread of wavelengths and often shows better off-axis behaviour (a
singlet asphere corrects spherical aberration on-axis but leaves coma
and field curvature). The recommendation: for a monochromatic on-axis
collimator the asphere is the right call; choose the achromat only if
the source is broadband, the field is wide, or small-quantity asphere
tooling cost dominates.
\end{exercise}

\begin{exercise}
A research group must choose between conventional optical microscopy
at NA = $1.4$ and super-resolution microscopy at $20\,\si{\nano
\meter}$ resolution for a study of cellular ultrastructure. Identify
the trade-offs in cost, sample preparation, time, and biological
relevance.
\paragraph{Solution sketch.} Conventional confocal at $\text{NA} = 1.4$
resolves about $220\,\si{\nano\meter}$, images live cells with standard
fluorophores, is fast, and is widely available; it cannot see structure
below the diffraction limit. Super-resolution at $20\,\si{\nano\meter}$
(STED, PALM/STORM) reaches the molecular scale but demands special
fluorophores or photoswitching, longer acquisition (especially
single-molecule localisation), higher cost and expertise, and often
fixed rather than live samples, raising the question of whether the
fixed-cell ultrastructure reflects the living state. The recommendation
turns on the science: if the target features are above
$\sim 250\,\si{\nano\meter}$ or live dynamics matter, conventional
microscopy serves; if the question is specifically sub-diffraction
ultrastructure and fixation is acceptable, the super-resolution cost is
justified.
\end{exercise}

\begin{exercise}
A research group must decide whether to deploy a $1\,\si{\meter}$
ground-based telescope with adaptive optics or to wait for time on
the JWST for an exoplanet-imaging programme. Identify the trade-offs
in resolution, sensitivity, schedule, and risk.
\paragraph{Solution sketch.} A $1\,\si{\meter}$ ground telescope with
adaptive optics reaches a diffraction limit of
$\sim 0.14$ arc-second in the near-infrared, is available on the group's
own schedule, and is cheap per night, but it loses sensitivity to sky
background and residual atmospheric correction, and high-contrast
exoplanet imaging is hard against the bright host star. The JWST offers
far greater sensitivity and stability and a clean point-spread function
from its $6.5\,\si{\meter}$ aperture, but time is scarce and
oversubscribed, the schedule is set by the proposal cycle, and the risk
is the wait and the competition. The recommendation: for bright,
nearby, wide-separation targets the ground telescope now is the
pragmatic choice; for faint or close-in companions where sensitivity and
stability decide, the JWST wait is warranted despite the schedule risk.
\end{exercise}
