% Volume 7: Information
% Opener: tagline + scope. Sits immediately after the \part{} declaration.

\vspace*{1.5cm}
\begin{center}
\itshape\large
Structure, representation, control, and constraint
\end{center}
\vspace{0.5cm}

\noindent
We cover logic, algorithms, software, hardware, data, communication, networks, and machine learning, treated as engineering. The reader leaves able to design and reason about computational systems with the same care a mechanical engineer applies to a load-bearing structure: assumptions named, failure modes identified, scale considered, ethical implications faced.

\vspace{0.5cm}

\section*{What this volume is for}

Volumes I through VI taught the engineer to measure, model, force, energise, build out of matter, and design with living systems. Each of those volumes treated its substrate as something that exists. The substrate of Volume VII is different. Information is not a thing the engineer encounters in the world; it is a structure the engineer imposes on the world to make the world legible, controllable, and shareable. A bit is not a particle. A program is not a machine. A protocol is not a wire. Yet bits, programs, and protocols are the working substance of the discipline whose products have, in seven decades, become the substrate of every other engineering discipline.

The reader leaves this volume able to do four things. First, reason about computational systems in the same vocabulary as physical systems: rate, accumulation, balance, scaling, failure mode, interface, transport. Second, build small but real software: a working algorithm, a usable program, a database the reader can query, a network connection the reader can trace, a learned model the reader can evaluate. Third, evaluate computational claims at engineering depth: what does a complexity bound actually say about a 10\,GB input, what does a 99\%-accurate classifier do on the case it gets wrong, what does a side-channel attack require of the attacker, what does a quantum speedup require of the hardware. Fourth, treat the failure of a software system at the same depth as the failure of a bridge or a reactor. The chapter on Volume X will name the discipline of failure; this volume installs the practice in advance.

The volume is one of the four dense volumes named in the project's settled questions (Q11). It is dense because computing has accumulated, in seven decades, a pile of distinct disciplines that share little except their reliance on logic and the historical accident that a computer is their substrate. Algorithms, systems, networks, security, machine learning, signal processing, and digital control are treated here as separable disciplines with shared archetypes, not as one undifferentiated thing called ``computing.''

\section*{What the reader brings}

The reader who opens Volume VII has finished Volumes I through VI and carries their habits. From Volume I: estimation before calculation, units through every step, error as a first-class quantity, the half-life tag, the failure section. From Volume II: derivatives, integrals, vectors and matrices, eigenvalues, probability, optimisation, numerical methods, discrete mathematics. From Volume III: the Newtonian discipline of forces and bodies, vibrations as second-order linear systems, the convection-diffusion equation. From Volume IV: energy as conserved currency, electromagnetism, circuits. From Volume V: materials and the molecular constraints they impose. From Volume VI: living systems as the existing prior art on information-driven engineering, four billion years deep.

Mathematically, the reader brings what Volume II delivered. Boolean algebra and the elementary propositional connectives sit at the start of Chapter 1; the reader has met them. Discrete mathematics (Chapter II.17) is the prerequisite for Chapters 1, 3, and 4. Probability and statistical inference (Chapters II.13, II.14) underpin Chapters 13, 17, and 18. Linear algebra and optimisation (Chapters II.9, II.10, II.15) underpin Chapters 17 and 18. Numerical analysis (II.16) underpins Chapters 2 and 7. Signal processing (Chapter 15) explicitly returns to the Fourier and sampling foundations of Chapters II.5 through II.7 and Chapter I.6, taking them through to filter design and modern audio-video coding.

We assume no specific prior programming experience. The reader who arrives knowing only what Volume II Chapter 17 has taught about discrete structures is the design point. We assume a willingness to install a working development environment, write code in at least two languages, run a database and a server on a personal machine, capture network traffic on a home network, and stand up a small cloud account. None of these is a credential; all of them are working habits.

\section*{The arc of the nineteen chapters}

The volume is organised in seven arcs.

\textbf{Arc one: foundations} (Chapters 1-2). Logic, sets, formal languages, computability, and number representation. Chapter 1 sets the formal language in which the rest of computing is described; Chapter 2 sets the actual representation of numbers in a finite-precision machine and inherits Volume I's discipline of accounting for error.

\textbf{Arc two: algorithms and data} (Chapters 3-4). Time and space complexity, the canonical algorithm families (greedy, divide-and-conquer, dynamic programming), graph algorithms, NP-completeness; arrays, hash tables, trees, heaps, graphs in memory, persistent structures, cache-aware layout. The two chapters are the engineer's working vocabulary for ``what runs in what time on what hardware.''

\textbf{Arc three: how computation is run} (Chapters 5-7). Programming languages and the paradigms they encode; operating systems as the engineer's interface to the machine; computer architecture from the digital-logic level through caches, branch prediction, vector and parallel execution. The arc closes at the side channel: the lesson that the abstraction of computation as logic does not survive contact with the physical substrate.

\textbf{Arc four: compilers and data} (Chapters 8-9). Compilers as the bridge from a high-level language to executable code; databases as the discipline of storing facts that survive change, concurrency, and the engineer's mistakes.

\textbf{Arc five: networks, distribution, security} (Chapters 10-12). The OSI-layered model and the actual protocols of the Internet; the hard problem of distributed systems where partial failure is the rule; cryptography and the engineering discipline that protects systems whose threat model is no longer an honest accidental fault.

\textbf{Arc six: information theory, quantum, signals, digital control} (Chapters 13-16). Shannon's framework as the foundation of every communication and compression claim; quantum information as the generalisation that classical information theory now contains; signal processing as the engineering machinery of Fourier analysis and sampling; control of digital systems as the bridge to Volume IX.

\textbf{Arc seven: machine learning and the practice of software} (Chapters 17-19). Machine learning fundamentals (regression, classification, generalisation, ensembles); modern AI (deep learning, transformers, large language models, multimodal models); software engineering as a discipline (testing, deployment, observability, the practices that make all the previous chapters reliable in production).

\section*{What the reader will be able to do at the end}

A reader who finishes Volume VII should be able to:

\begin{itemize}[leftmargin=*]
  \item Write a clear propositional or first-order predicate
    specification of an engineering requirement, and demonstrate
    that two reasonable readers would interpret it the same way.
  \item Identify the precision of a numerical computation, account
    for its accumulated rounding and truncation error, and
    recognise the cases where IEEE 754 semantics produce results
    surprising to non-numerical engineers.
  \item Implement and benchmark sorting, searching, and graph
    algorithms; reason about their time and space complexity at
    the engineering level; identify the input regimes in which an
    $\oom{n^{2}}$ algorithm is preferable to an $\oom{n \log n}$
    one.
  \item Choose an appropriate data structure for a given workload
    and explain the trade-off against the alternatives.
  \item Read and write programs in at least two languages from
    different paradigms (imperative, functional, declarative), and
    explain the design forces that pushed each language into the
    shape it has.
  \item Reason about a running operating system: trace what happens
    in a system call, identify when paging will thrash, recognise
    a scheduling problem, name three places interrupts can be
    lost.
  \item Read an instruction-set manual, profile a tight loop on a
    real machine, and explain the observed performance via cache
    behaviour and branch prediction.
  \item Design a relational schema for a non-trivial domain; write
    queries that exploit indexes; reason about the isolation level
    a workload actually needs.
  \item Trace a HTTPS connection end to end at the protocol level
    and identify at each layer what is doing the work.
  \item Identify partial-failure modes in a distributed system and
    name the consistency model a workload needs.
  \item Design a small authenticated, integrity-protected protocol
    and write the threat model it survives.
  \item Compute entropies, mutual informations, and the capacity of
    a simple channel; design and implement a Huffman code; explain
    what an error-correcting code does in the presence of noise.
  \item Design a digital filter from a frequency specification and
    characterise its response.
  \item Build a working machine-learning classifier from scratch,
    diagnose its failure modes, and recognise when the available
    data does not support the claimed conclusion.
  \item Fine-tune a small open language model on a domain task and
    evaluate it without fooling oneself.
  \item Ship a small project end-to-end with version control,
    tests, CI, deployment, and a working observability stack.
\end{itemize}

A reader who reaches Volume VIII without these capabilities has read Volume VII without doing it. We say so plainly because Volume VIII's electronics chapters assume the architecture of this volume's Chapter 7, and Volume IX's control chapters assume the digital control of this volume's Chapter 16.

\section*{Archetypes introduced or developed in this volume}

The archetypes of Volume VII are familiar; what changes is the substrate. Scaling has new units (cycles per second, bits per second, gigabytes per second). Balance now includes the consistency-availability balance of distributed systems. Failure now includes the side channel, the silent data corruption, the off-by-one, the unspecified case. Interface now includes the application programming interface, the protocol stack, and the unit-test boundary. Transport now includes the network packet. Uncertainty now includes the random-access memory bit-flip, the floating-point round-off, the model that is wrong outside its training distribution. Control gets its discrete-time formal home in Chapter 16. Optimisation appears in algorithm design (Chapter 3), query planning (Chapter 9), compiler optimisation (Chapter 8), machine learning (Chapters 17-18), and signal processing (Chapter 15).

Two archetypes that Volume I did not have are first installed here. \engterm{Concurrency} appears in operating systems (Chapter 6), languages (Chapter 5), databases (Chapter 9), distributed systems (Chapter 11), and digital control (Chapter 16). \engterm{Ethics under capability} appears in cryptography (Chapter 12), machine learning (Chapter 17), modern AI (Chapter 18), and software engineering (Chapter 19), and is fully developed in Volume XI.

\section*{The volume project}

The reader builds, deploys, and operates a small but real software system: a personal index of the measurement notebooks from Volumes I through VI, with a programmable API, persistent storage, authentication, monitoring, automated tests, and a deployment pipeline. The system survives a hostile review for security, reliability, scalability, and ethical handling of data. The review is the credential by which the reader earns admission to Volume VIII.

The project is described in detail across the chapters that contribute to it. Chapter 4 picks the data structure; Chapter 5 picks the languages; Chapter 9 specifies the schema; Chapter 10 carries the protocol design; Chapter 12 picks the authentication and the threat model; Chapter 19 carries the test, deployment, and observability discipline.

\section*{A note on the next volume}

Volume VIII takes up Machines: mechanisms, machine elements, electronics, sensors, actuators, robotics. The bridge is direct. The reader who has built a small distributed software system and reasoned about its failure modes is ready to apply the same discipline to a system that has, in addition to the bits, the masses, the forces, and the energy of physical engineering. The information substrate of Volume VII is what makes the machines of Volume VIII controllable; the physical substrate of Volume VIII is what makes the information of Volume VII consequential.

\vspace{1cm}
