\chapter{Plates, shells, solid mechanics}
\label{vol03:ch09}

\epigraph{The reader is reminded that the bending theory of plates is
based on the same assumptions as the elementary bending theory of
beams, with the additional assumption that the deflections are small
in comparison with the thickness of the plate.}{Stephen Timoshenko
and S.\ Woinowsky-Krieger, \emph{Theory of Plates and Shells}, 2nd
ed.\ (1959), preface to chapter 1
\cite[p.~v]{text:timoshenko-plates-shells}.}

\chapmeta{Half-life: durable for plate and shell theory; medium for finite-element practice. Archetypes: stability, scaling, failure. Exercise target: 35. Project track: analytical plus simulation (hybrid). Hazard class: none. Reader path default: standard, with sections explicitly tagged. Named cases: Hubble Space Telescope primary mirror (1990, cross-referenced for thin-mirror figure error and the metrology-chain failure pattern); de Havilland Comet (1954, invoked for fatigue crack initiation at square-cornered cabin cutouts, the structural counterpart of the stress-concentration discipline).}

\noindent
A beam is a one-dimensional structural element; a plate is the
two-dimensional generalisation, a shell is the curved two-dimensional
generalisation, and three-dimensional elasticity is the parent theory
beneath all three. The chapter develops Kirchhoff-Love plate theory
in the small-deflection limit, the membrane and bending theory of
thin shells, the equations of three-dimensional linear elasticity in
the form the finite-element method (FEM) uses them, the working
discipline of mesh generation and convergence checking, and the
verification, validation, and uncertainty quantification that turn
an FEM output into a load-bearing engineering claim. The failure
section names the mechanisms by which simulation results mislead
engineers who trust them without verifying them.

\section{Plate theory}\pathtag{core}

A \engterm{plate} is a structural element whose thickness is small
compared to its in-plane dimensions and which carries transverse load
primarily by bending. The plate analogue of the Euler-Bernoulli beam
theory of chapter 8 is the Kirchhoff-Love plate theory, due to
Gustav Kirchhoff (1850) and later refined by A.\ E.\ H.\ Love. The
theory rests on three kinematic assumptions, parallel to those of
slender-beam theory:

\begin{itemize}[leftmargin=*]
  \item Straight lines normal to the midsurface before deformation
    remain straight and normal after deformation.
  \item These normals do not change length: the plate's thickness is
    preserved.
  \item The transverse normal stress $\sigma_{zz}$ through the
    thickness is small compared to the in-plane stresses.
\end{itemize}

The first two assumptions, the \engterm{Kirchhoff hypothesis},
neglect transverse shear deformation and are accurate when the
plate's thickness $h$ is small compared to its smallest in-plane
dimension (the working rule is $h \le L/20$). For thicker plates
the Mindlin-Reissner theory (section 9.3) keeps transverse shear and
gives more accurate results.

Under the Kirchhoff hypothesis, the displacement field of a plate
with midsurface deflection $w(x, y)$ is
\[
u(x, y, z) = -z \frac{\partial w}{\partial x}, \quad
v(x, y, z) = -z \frac{\partial w}{\partial y}, \quad
w(x, y, z) = w(x, y),
\]
where $z$ is measured from the midsurface and $(u, v, w)$ are the
displacement components in the $(x, y, z)$ directions. The strains
follow:
\[
\epsilon_{xx} = -z \frac{\partial^2 w}{\partial x^2}, \quad
\epsilon_{yy} = -z \frac{\partial^2 w}{\partial y^2}, \quad
\gamma_{xy} = -2 z \frac{\partial^2 w}{\partial x \partial y},
\]
linear in $z$, as in beams. The in-plane stresses, by Hooke's law
for an isotropic plate in plane stress (Young's modulus $E$,
Poisson's ratio $\nu$), are
\begin{align*}
\sigma_{xx} &= \frac{E}{1 - \nu^2}\left(\epsilon_{xx} + \nu
\epsilon_{yy}\right), \\
\sigma_{yy} &= \frac{E}{1 - \nu^2}\left(\epsilon_{yy} + \nu
\epsilon_{xx}\right), \\
\tau_{xy} &= \frac{E}{2(1+\nu)} \gamma_{xy}.
\end{align*}

The bending moments per unit length of the plate's midsurface are
the first moments of the stresses through the thickness,
\[
M_{xx} = \int_{-h/2}^{h/2} z \sigma_{xx} \,dz, \quad
M_{yy} = \int_{-h/2}^{h/2} z \sigma_{yy} \,dz, \quad
M_{xy} = \int_{-h/2}^{h/2} z \tau_{xy} \,dz.
\]
Performing the integrals,
\[
M_{xx} = -D \left(\frac{\partial^2 w}{\partial x^2} + \nu
\frac{\partial^2 w}{\partial y^2}\right), \quad
M_{yy} = -D \left(\frac{\partial^2 w}{\partial y^2} + \nu
\frac{\partial^2 w}{\partial x^2}\right),
\]
\[
M_{xy} = -D(1-\nu) \frac{\partial^2 w}{\partial x \partial y},
\]
where the constant
\[
D = \frac{E h^3}{12(1 - \nu^2)}
\]
is the plate's \engterm{flexural rigidity}, the two-dimensional
analogue of the beam's $EI$. The cube on thickness is structurally
identical to the cube on depth in the beam: a plate twice as thick
has eight times the flexural rigidity.

Transverse force equilibrium of an infinitesimal plate element,
combined with the moment-curvature relations above, gives the
\engterm{plate bending equation}, the two-dimensional analogue of
the Euler-Bernoulli beam equation $EI \,y'''' = w(x)$:
\[
D \nabla^4 w = q(x, y),
\]
with $\nabla^4 = \nabla^2 \nabla^2$ the biharmonic operator and
$q(x, y)$ the distributed transverse load per unit area. The
equation is a fourth-order linear PDE in two spatial variables; the
boundary conditions on the plate's edge (two conditions per edge)
close the problem.

The route from the moment-curvature relations to the biharmonic is
worth walking once in full, because the same elimination recurs for the
shell and reappears in the derivation exercises. Transverse force
balance on a $dx\,dy$ element, with $Q_x$ and $Q_y$ the transverse
shear forces per unit length, reads
\[
\frac{\partial Q_x}{\partial x} + \frac{\partial Q_y}{\partial y}
+ q = 0 .
\]
Moment balance about the two in-plane axes relates the shears to the
gradients of the moment resultants,
\[
Q_x = \frac{\partial M_{xx}}{\partial x}
+ \frac{\partial M_{xy}}{\partial y}, \qquad
Q_y = \frac{\partial M_{yy}}{\partial y}
+ \frac{\partial M_{xy}}{\partial x},
\]
so substituting the shears out of the force balance leaves a single
equation in the moments,
\[
\frac{\partial^2 M_{xx}}{\partial x^2}
+ 2\frac{\partial^2 M_{xy}}{\partial x \partial y}
+ \frac{\partial^2 M_{yy}}{\partial y^2}
= -q .
\]
Now insert the moment-curvature relations. The first term carries
$-D(w_{,xxxx} + \nu\, w_{,xxyy})$, the third carries
$-D(w_{,yyyy} + \nu\, w_{,xxyy})$, and the twist term carries
$-2D(1-\nu)\, w_{,xxyy}$. Adding them, the two $\nu w_{,xxyy}$
contributions and the $-2D(1-\nu) w_{,xxyy}$ contribution combine to
the clean coefficient $2$ on the mixed fourth derivative, because
$2\nu + 2(1-\nu) = 2$, and the Poisson ratio drops out of the
governing operator entirely:
\[
D\left(\frac{\partial^4 w}{\partial x^4}
+ 2\frac{\partial^4 w}{\partial x^2 \partial y^2}
+ \frac{\partial^4 w}{\partial y^4}\right) = q .
\]
The bracket is exactly $\nabla^4 w$. That $\nu$ survives in the
moments and in the recovered stresses but cancels out of the
deflection operator is the reason the centre deflection of a plate
depends on Poisson's ratio only weakly through $D$, while the edge
moment of a clamped plate carries $\nu$ directly.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-plate-element}

The moment resultants of figure~\ref{fig:vol03ch09:plate-element} are
the bookkeeping device that turns the three-dimensional stress field
into a two-dimensional field of stress couples per unit length. Once
$w(x,y)$ is known, the through-thickness stress at any depth $z$
recovers from the moments by the linear distribution
\[
\sigma_{xx}(x,y,z) = \frac{12 M_{xx}}{h^3}\, z, \qquad
\sigma_{xx}^{\max} = \pm \frac{6 M_{xx}}{h^2}\ \text{at}\ z = \pm h/2,
\]
the plate analogue of the beam flexure formula $\sigma = Mc/I$ with
$c = h/2$ and $I = h^3/12$ per unit width. The surface stress, not the
midsurface stress (which is zero in pure bending), is the quantity to
compare against the material's allowable.

The standard plate boundary conditions, on an edge $x = a$:

\begin{itemize}[leftmargin=*]
  \item Simply supported (pin or roller along the edge):
    $w = 0$, $M_{xx} = 0$. The plate's edge deflection vanishes;
    the moment normal to the edge vanishes.
  \item Clamped (built-in along the edge): $w = 0$, $\partial w/
    \partial x = 0$. The deflection and the slope normal to the
    edge both vanish.
  \item Free: $M_{xx} = 0$ and the Kirchhoff effective shear $V_{xx}
    = \partial M_{xx}/\partial x + 2 \partial M_{xy}/\partial y = 0$.
    The effective shear combines the transverse shear with the
    twisting moment's gradient because the Kirchhoff hypothesis does
    not admit independent twist; see \cite[ch.~2]{
    text:timoshenko-plates-shells} for the derivation.
\end{itemize}

The free edge deserves a second look, because its two conditions hide a
subtlety that the beam theory has no analogue for and that a careless analyst
mishandles. A beam edge carries at most two force quantities, a shear and a
moment, and a free beam end sets both to zero. A plate edge carries three: a
bending moment $M_{xx}$ normal to the edge, a transverse shear $Q_x$, and a
twisting moment $M_{xy}$ along the edge. A fourth-order equation admits only
two boundary conditions per edge, so three quantities cannot each be set to
zero independently. Kirchhoff resolved the count by observing that the
twisting moment along the edge is statically equivalent to a distribution of
transverse shear: a twisting couple $M_{xy}$ over an edge length $dy$ can be
replaced by two opposite vertical forces $M_{xy}$ a distance $dy$ apart, and
where the twist varies along the edge these paired forces leave a net
distributed shear $\partial M_{xy}/\partial y$. Adding it to the true
transverse shear gives the \engterm{Kirchhoff effective shear} $V_{xx} =
Q_x + \partial M_{xy}/\partial y = \partial M_{xx}/\partial x + 2\,\partial
M_{xy}/\partial y$, and the free edge sets $M_{xx} = 0$ and $V_{xx} = 0$, two
conditions, matching the order of the equation. The replacement leaves a
residue at any corner where two edges meet: the paired forces from the two
edges do not cancel there but sum to a concentrated \engterm{corner force}
$R = 2 M_{xy}$. A simply supported rectangular plate under downward load
develops upward corner forces, which is to say the corners of the plate try to
lift off their supports, and a plate merely resting on its corners without
hold-downs will rise there and redistribute its load. The effect is real and
measurable: an uplifted corner carries none of the load the analysis assigned
it, and a plate held down at the corners carries a different moment field from
one free to lift. The corner force is the plate-theory consequence of the
Kirchhoff hypothesis's refusal to admit an independent twist, and it is the
reason a square simply supported plate is often detailed with anchors at its
corners.

\engterm{Closed-form solutions} exist for a handful of canonical
cases. A simply supported rectangular plate of dimensions $a \times
b$ under uniform load $q_0$ has the Navier double-sine series
solution
\[
w(x, y) = \frac{16 q_0}{\pi^6 D} \sum_{m, n \text{ odd}}
\frac{\sin(m\pi x/a) \sin(n\pi y/b)}{m n \left[(m/a)^2 + (n/b)^2
\right]^2}.
\]
The series converges rapidly; the first term ($m = n = 1$) typically
gives the centre deflection to within a few percent. For a square
plate of side $a$, $w_{\max} \approx 0.00406 q_0 a^4/D$, a useful
working number.

A clamped circular plate of radius $a$ under uniform load $q_0$ has
the axisymmetric closed-form solution
\[
w(r) = \frac{q_0 (a^2 - r^2)^2}{64 D}, \quad
w_{\max} = w(0) = \frac{q_0 a^4}{64 D},
\]
the standard reference case for circular-plate analysis (Timoshenko
\cite[ch.~3, eqn.~62]{text:timoshenko-plates-shells}). The
maximum bending moment occurs at the clamped edge, where
$M_{rr} = -q_0 a^2/8$.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-circular-plate-deflection}

Figure~\ref{fig:vol03ch09:circular-plate-deflection} contrasts the
clamped and simply supported deflection shapes. The simply supported
circular plate carries the same load with about four times the centre
deflection, because its edge rotates freely instead of being held flat.
The clamped plate pays for its smaller deflection with a large edge
moment, which is why a clamped circular plate of brittle material
tends to crack at the rim, while a simply supported one of the same
material tends to crack at the centre. The collected deflection and
moment coefficients for the canonical plate cases are tabulated in
\repopath{data/plate_coefficients.csv} for quick design reference.

The circular-plate solution is reached by collapsing the biharmonic to
one variable, the move that turns a PDE into an ODE and makes the
closed forms above derivable by hand. For an axisymmetric plate the
deflection depends on the radius alone, $w = w(r)$, and the Laplacian
in polar coordinates loses its angular term, $\nabla^2 = d^2/dr^2 +
(1/r)\,d/dr$. The plate equation becomes the fourth-order ordinary
equation
\[
\nabla^4 w = \frac{1}{r}\frac{d}{dr}\!\left[r\,\frac{d}{dr}\!\left(
\frac{1}{r}\frac{d}{dr}\!\left(r\,\frac{dw}{dr}\right)\right)\right]
= \frac{q}{D} ,
\]
which integrates four times in closed form. For a uniform load $q_0$
the particular solution is $w_p = q_0 r^4/(64 D)$, and the homogeneous
solution adds $C_1 + C_2 r^2 + C_3 \ln r + C_4 r^2 \ln r$. A solid
plate (no central hole) cannot carry the singular $\ln r$ and
$r^2 \ln r$ terms, so $C_3 = C_4 = 0$ and only $C_1$ and $C_2$ survive,
fixed by the two edge conditions. For the clamped edge, $w(a) = 0$ and
$w'(a) = 0$ give $C_2 = -q_0 a^2/(32 D)$ and $C_1 = q_0 a^4/(64 D)$,
which reassemble to the catalogue form
\[
w(r) = \frac{q_0\,(a^2 - r^2)^2}{64 D},
\]
the result quoted above, now derived rather than asserted. The radial
and circumferential moments follow from the axisymmetric
moment-curvature relations $M_r = -D(w'' + \nu w'/r)$ and
$M_\theta = -D(w'/r + \nu w'')$. They are plotted in
figure~\ref{fig:vol03ch09:axisymmetric-plate}: the radial moment is
sagging at the centre and swings to its largest magnitude, the hogging
$-q_0 a^2/8$, at the clamped rim, while the circumferential moment
stays smaller and positive. The rim governs the clamped plate, which is
the analytic statement of the cracking pattern noted above.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-axisymmetric-plate}

For non-canonical geometries, the working tool is the finite-element
method (section 9.5); the closed-form solutions remain the reference
against which the FEM result is verified.

\begin{principle}[Plates carry transverse load with thickness cubed]
For a thin isotropic plate of thickness $h$ in pure bending, the
flexural rigidity scales as $D = E h^3/[12(1-\nu^2)]$. The thickness
appears cubed because the bending moment per unit length is the
first moment of the linearly varying stress through the thickness,
and the integral picks up an additional factor of $h$ from the
moment arm. Doubling the plate's thickness multiplies its bending
stiffness by eight; halving the thickness divides the stiffness by
eight. The thickness is therefore the most powerful parameter in
plate design, exactly as the depth is the most powerful parameter
in beam design.
\end{principle}

\section{Shell structures}\pathtag{standard}

A \engterm{shell} is a curved plate: a thin structural element whose
midsurface is a curved surface and which carries load by both
bending and membrane action. The fundamental difference between a
plate and a shell is that a shell can carry transverse load by
in-plane (membrane) stresses, with bending playing only a secondary
role. An eggshell carries the weight of a hen by membrane
compression; a folded paper carries far more transverse load than a
flat paper because the folds give it shell-like geometry.

For a thin shell of midsurface radius $R$ and thickness $h$, with
$h/R \ll 1$, the stress state at any point is decomposed into
membrane stresses (independent of $z$ through the thickness, in
equilibrium with the applied load) and bending stresses (linear in
$z$, vanishing on the midsurface). The decomposition is exact for
the linear-elastic shell.

\engterm{Membrane theory} treats the shell as a curved surface that
carries only in-plane stress resultants, $N_{xx}, N_{yy}, N_{xy}$
(forces per unit length of midsurface), with no bending. For a
shell of revolution under axisymmetric load (a pressure vessel, a
water tower, a dome), membrane theory gives the meridional and
circumferential stress resultants directly from force equilibrium
on a strip of the shell.

The two governing equations are worth setting down once in their
general form, because every closed result in this section is a special
case of them. Take a shell of revolution loaded by a normal pressure
$p$, with meridional resultant $N_\phi$ (per unit length, along the
generating curve) and circumferential resultant $N_\theta$ (per unit
length, around the parallel circle), and let $r_1$ and $r_2$ be the two
principal radii of curvature at a point, $r_1$ in the meridian plane
and $r_2$ measured along the surface normal out to the axis
(figure~\ref{fig:vol03ch09:shell-revolution-element}). Normal
equilibrium of an element, the balance of the pressure against the
inward components of the two resultants curved over their respective
radii, gives the \engterm{Laplace-Young equation}
\[
\frac{N_\phi}{r_1} + \frac{N_\theta}{r_2} = p .
\]
This single relation already shows why curvature is what lets a shell
carry transverse load without bending: each resultant contributes an
inward force proportional to its own curvature, and a flat sheet
($r_1, r_2 \to \infty$) contributes nothing, which is why a flat plate
must bend. The second equation comes from axial force balance on the
cap of the shell below a parallel circle of radius $\rho = r_2\sin\phi$.
The vertical resultant of the pressure on that cap is carried by the
vertical component of $N_\phi$ acting around the circle, so
\[
2\pi \rho\, N_\phi \sin\phi = (\text{axial load carried below } \phi),
\]
which fixes $N_\phi$ directly from the load, after which the
Laplace-Young equation returns $N_\theta$. The membrane problem is in
this sense statically determinate: two equilibrium equations give the
two resultants with no appeal to the material law at all, the property
that makes membrane sizing the first calculation in any vessel design.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-shell-revolution-element}

The standard examples drop straight out of the two equations. For a
sphere both radii equal $R$, the symmetry forces $N_\phi = N_\theta = N$,
and Laplace-Young gives $2N/R = p$, hence $N = pR/2$ and the wall stress
$\sigma = N/h = pR/2h$. For a cylinder the meridian is straight, so
$r_1 \to \infty$ and the meridional term vanishes; the equation reduces
to $N_\theta/R = p$, the hoop resultant $N_\theta = pR$ and stress
$pR/h$, while the axial resultant comes from the end-cap balance,
$N_z = pR/2$ and stress $pR/2h$. The cone follows the same way with the
slant geometry inserted. The three results below are not three formulas
to memorise but one master equation read at three geometries:

\begin{itemize}[leftmargin=*]
  \item \engterm{Thin spherical pressure vessel} of radius $R$,
    thickness $h$, internal gauge pressure $p$:
    \[
    \sigma_{\text{meridional}} = \sigma_{\text{circumferential}} =
    \frac{pR}{2h}.
    \]
    Both principal membrane stresses are equal; the sphere is the
    most efficient pressure-containing geometry.
  \item \engterm{Thin cylindrical pressure vessel} of radius $R$,
    thickness $h$, length $L \gg R$, internal gauge pressure $p$:
    \[
    \sigma_{\text{circumferential (hoop)}} = \frac{pR}{h}, \quad
    \sigma_{\text{axial}} = \frac{pR}{2h}.
    \]
    The hoop stress is twice the axial stress; the cylinder is
    twice as efficient in axial as in hoop, and this is why
    pressure-vessel design uses the hoop stress as the governing
    quantity.
  \item \engterm{Conical shell} of half-angle $\alpha$ under
    internal pressure $p$, at distance $s$ along the slant from the
    apex:
    \[
    \sigma_{\text{circumferential}} = \frac{p s \tan\alpha}{h},
    \quad \sigma_{\text{meridional}} = \frac{p s \tan\alpha}{2 h}.
    \]
    The stresses grow linearly with distance from the apex; the
    cone is therefore most heavily stressed at its base.
\end{itemize}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-cylinder-stresses}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-vessel-efficiency}

The three membrane results above share a single algebraic skeleton:
the membrane stress is the product of the pressure, a length scale set
by the local radius of curvature, and the reciprocal of the wall
thickness. The sphere splits the load between two equal principal
curvatures and so carries it at half the stress of the cylinder, whose
hoop direction has finite curvature while the axial direction is
straight. Figure~\ref{fig:vol03ch09:vessel-efficiency} turns this into
the design statement an engineer uses: for the same radius and
allowable stress, a sphere needs half the wall thickness of a
cylinder. The membrane sizing and its inversion to a required wall
thickness are implemented in \repopath{code/pressure_vessel.py}.

The membrane-theory result is exact within the membrane hypothesis
(no bending, no transverse shear, no thickness through-stress) and
is accurate for thin shells away from edge constraints, point loads,
or geometric discontinuities. Near such locations, bending stresses
that are not small compared to the membrane stresses arise as a
\engterm{boundary layer} of bending in the shell. The bending decays
away from the discontinuity over a length of order $\sqrt{Rh}$, narrow
for a thin shell, and inside that band it can reach a sizeable fraction
of the membrane stress (figure~\ref{fig:vol03ch09:discontinuity-stress};
the governing equation and its damped-oscillator structure are derived
in the mastery box below). Every welded junction, supported edge, and
nozzle in a real vessel carries this discontinuity stress, which is why
the membrane sizing of a vessel is necessary but never sufficient. The
working tool for the full boundary-layer analysis is the
Donnell-Mushtari-Vlasov equation set, beyond the scope of this chapter
(see \cite[ch.~14]{text:timoshenko-plates-shells}).

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-discontinuity-stress}

\engterm{Shell buckling}. A thin shell in compression buckles at a
load far below the equivalent flat-plate buckling load and far below
the prediction of linear shell-buckling theory. The classical Euler
buckling result for a thin spherical shell of radius $R$ under
external pressure is
\[
p_{\text{cr}} = \frac{2 E h^2}{R^2 \sqrt{3(1-\nu^2)}},
\]
but experiment shows that real spherical shells buckle at $20$ to
$40\,\%$ of this value because of geometric imperfections. A small
flat spot, a small dimple, or a small out-of-roundness reduces the
buckling load substantially. The companion result for a long
cylindrical shell under uniform external pressure, the case of a
submarine hull or a vacuum-loaded tank between widely spaced
stiffeners, is the ring-collapse formula
\[
p_{\text{cr}} = \frac{E h^3}{4(1-\nu^2) R^3},
\]
which follows from treating a unit-length ring of the cylinder as an
Euler ring and which falls off as the cube of the thinness ratio
$h/R$. A cylinder twice as thin (at fixed radius) loses a factor of
eight in external-pressure capacity, the same thickness-cubed scaling
that governs plate bending. The same imperfection knock-down applies:
the working capacity is a fraction of this linear value, and the
design adds ring stiffeners to shorten the unsupported length when the
bare shell cannot carry the external pressure (worked example in the
exercises, and the external-pressure step of the air-receiver case
study below). The shell-buckling literature calls
this the \engterm{imperfection sensitivity} of shells, and it is the
single most important practical issue in shell design. Design
codes for shells (ASME Boiler and Pressure Vessel Code, Section VIII)
use reduced buckling capacities calibrated to extensive test data;
no engineer designs a shell against its linear-theory buckling load.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-shell-buckling}

The contrast with the flat plate is the instructive one.
A flat plate in compression buckles close to its linear-theory load,
within ten or twenty percent, because the post-buckling path is stable
and small imperfections only nudge the result. A shell buckles on an
unstable post-buckling path, where the first dimple relieves the
membrane stress that was holding the neighbouring material, which
dimples in turn. The result is the steep knock-down of
figure~\ref{fig:vol03ch09:shell-buckling} and the worked design
factors collected in \repopath{data/shell_knockdown.csv}: the axially
compressed cylinder, the most sensitive case, is designed to roughly
one fifth of its classical load.

The difference between the stable plate and the unstable shell is not a matter
of degree but of the shape of the load-deflection curve just past the critical
point, and the general theory of it is due to Koiter. Track the applied load
against the amplitude of the buckle. For the flat plate the curve rises after
buckling: the plate goes on carrying more load as the buckle grows, the
post-buckling reserve that worked example 32 will bank on, and this rising
path is why a small imperfection only rounds the corner of the curve and
barely lowers the peak. For the axially compressed cylinder the curve falls
after the critical point: once a dimple forms it takes less load, not more, to
deepen it, because the dimpled material has shed the membrane compression that
was bracing its neighbours, which dimple in their turn. A structure on a
falling post-buckling path cannot sit at the classical load at all in the
presence of any imperfection, because the imperfection lets the collapse begin
early and the falling path carries it away. Koiter's analysis makes the
sensitivity quantitative: for the worst shell cases the collapse load falls
below the classical value in proportion to the two-thirds power of the
imperfection amplitude, so an out-of-roundness of only a few percent of the
wall thickness pulls the capacity down by half, and there is no threshold
below which the imperfection stops mattering. That two-thirds-power law, not a
safety-factor tradition, is the reason shell knock-down factors are as severe
as they are and why they are keyed to a measured or specified geometric
tolerance rather than to a load uncertainty. The plate tolerates its
imperfections because its post-buckling path forgives them; the shell punishes
its imperfections because its path amplifies them, and the whole discipline of
shell design follows from which way that curve bends.

\engterm{The Brazier effect}. A thin-walled tube under bending
ovalises: the originally circular cross-section flattens into an
ellipse at the section of maximum bending moment, because the
compression-side fibres push inward while the tension-side fibres
pull outward. The ovalisation reduces the section's effective
bending stiffness and triggers a nonlinear failure mode at a moment
significantly below the linear prediction. Brazier (1927) gave the
classical analysis; modern thin-tube design (bicycle frames, aircraft
fuselages, oil pipelines) accounts for the Brazier effect in the
design rules for tube wall thickness.

The mechanism is a feedback loop, which is what makes it a stability
problem rather than a stress problem. Bending the tube puts the longitudinal
fibres above the neutral axis in tension and those below in compression, and
each curved fibre, held to the tube's curvature, presses toward the tube
axis with a small radial component. Those radial components are unbalanced
around the section, and their net effect flattens the circle toward an
ellipse. The flattening moves material toward the neutral axis, which
lowers the section's second moment of area, which for a given bending moment
raises the fibre stresses and the curvature, which flattens the section
further. The loop runs away at a limit moment, the Brazier moment, above
which no equilibrium of the round tube exists and the section collapses by
ovalisation, typically at about two-thirds of the moment that linear
beam-on-tube theory would predict for first yield. A separate and usually
lower limit is local buckling of the compression side into short-wavelength
ripples, the axial-compression shell buckling of the previous paragraph
riding on top of the ovalisation. A drinking straw bent between the fingers
shows both: it flattens smoothly (Brazier) and then kinks sharply at one
section (local buckling), and once kinked it carries almost no moment. The
design responses are to limit the diameter-to-thickness ratio so the Brazier
and local-buckling moments stay above the working moment, to fill or
stiffen the tube where a high moment is unavoidable (the mandrel in a
tube-bending machine exists to stop exactly this ovalisation), and, in a
pipeline laid over a seabed, to limit the curvature the pipe is allowed to
take during installation.

\begin{mastery}
\textbf{The shell bending boundary layer.} Membrane theory satisfies
force equilibrium but cannot, on its own, satisfy the displacement
conditions at an edge, a support, or a junction between two shells of
different stiffness. A spherical head welded to a cylindrical shell is
the standard case: under pressure the two free shells want to grow
radially by different amounts (worked example 8), and the weld forces
them to share an edge. The shell reconciles the mismatch by developing
a self-equilibrating set of edge shears and moments that decay away
from the discontinuity. Linearising the axisymmetric shell equations
about the membrane state gives, for the radial displacement
perturbation $w$, an equation of the damped-oscillator form
\[
\frac{d^4 w}{dx^4} + 4\beta^4 w = 0, \qquad
\beta^4 = \frac{3(1-\nu^2)}{R^2 h^2},
\]
whose decaying solutions are $w(x) = e^{-\beta x}(C_1\cos\beta x +
C_2\sin\beta x)$. The structure is identical to the beam on an elastic
foundation of chapter 8: the membrane stiffness of the shell plays the
role of the foundation, and the bending stiffness plays the role of the
beam. The decay length $1/\beta$ scales as $\sqrt{R h}$, so the bending
boundary layer is narrow for a thin shell, a few times $\sqrt{Rh}$
wide. Inside that band the bending stress can reach a sizeable fraction
of the membrane stress and must be added to it; outside it the membrane
solution is accurate. The practical consequences run through every
pressure-vessel detail: the knuckle radius of a torispherical head, the
thickness taper at a nozzle, the reinforcing ring at a support saddle,
each exists to keep the boundary-layer stress within the allowable. The
same mathematics governs the edge of a long cylindrical tank under
hydrostatic load and the junction of a chimney to its foundation. A
finite-element model resolves the boundary layer only if the mesh
places several elements within one decay length of the discontinuity;
a mesh coarser than $\sqrt{Rh}$ at a junction reports the membrane
stress and misses the very stress that governs the detail.
\end{mastery}

\section{Three-dimensional elasticity and the FEM equations}\pathtag{standard}

The Kirchhoff plate theory and the membrane shell theory are
\engterm{reduced models}: they collapse one or more spatial
dimensions by imposing kinematic assumptions on the displacement
field. The full \engterm{three-dimensional linear elasticity}
equations make no such assumption; they govern the displacement
field $\vec{u}(x, y, z) = u\hat{\imath} + v\hat{\jmath} + w\hat{k}$
of an arbitrary solid body under load.

The three governing equation sets, in compact tensor notation:

\engterm{Kinematics} (strain-displacement, infinitesimal-strain
limit):
\[
\epsilon_{ij} = \tfrac{1}{2}\left(\frac{\partial u_i}{\partial x_j}
+ \frac{\partial u_j}{\partial x_i}\right).
\]
The strain tensor $\epsilon_{ij}$ is symmetric ($\epsilon_{ij} =
\epsilon_{ji}$) and has six independent components in three
dimensions; the relation to displacement is the same kinematic
identity introduced in chapter 3 of this volume.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-stress-element-3d}

The companion object is the stress tensor of
figure~\ref{fig:vol03ch09:stress-element-3d}, also symmetric and also
six-component. The symmetry $\sigma_{ij} = \sigma_{ji}$ is not an
assumption; it is the statement of moment equilibrium of the
infinitesimal element, an unbalanced shear couple would give the
element an infinite angular acceleration. Six strains, six stresses,
three displacements: these fifteen fields at every point of the body,
tied by the six kinematic relations, the six constitutive relations,
and the three equilibrium equations, are exactly the unknowns and
equations the finite-element method discretises.

\engterm{Constitutive law} (Hooke's law, isotropic linear-elastic
material):
\[
\sigma_{ij} = \lambda \delta_{ij} \epsilon_{kk} + 2 \mu
\epsilon_{ij},
\]
with $\lambda, \mu$ the Lam{\'e} parameters ($\mu$ also the shear
modulus $G$, and $\lambda + 2\mu/3 = K$ the bulk modulus). The
material constants are two: any pair of $(E, \nu)$, $(E, G)$,
$(K, G)$, or $(\lambda, \mu)$ uniquely determines the others by
algebraic identities. Representative values of $E$, $\nu$, $G$, yield
strength, and density for the materials this chapter uses are
collected in \repopath{data/elastic_constants.csv}.

\engterm{Equilibrium} (Cauchy stress, no body force besides
gravity):
\[
\frac{\partial \sigma_{ij}}{\partial x_j} + \rho g_i = 0,
\]
with $\rho$ the mass density and $g_i$ the gravitational
acceleration. The equation is the three-dimensional analogue of
$dM/dx = V$, $dV/dx = -w$ in beams: equilibrium of an
infinitesimal element.

The boundary conditions on a body of volume $\Omega$ with boundary
$\partial\Omega$ are imposed in two parts:

\begin{itemize}[leftmargin=*]
  \item \engterm{Displacement (Dirichlet) boundary conditions} on
    $\Gamma_u \subset \partial\Omega$: $\vec{u} = \vec{u}_0$ on
    $\Gamma_u$. The body is held in place along that part of the
    boundary.
  \item \engterm{Traction (Neumann) boundary conditions} on
    $\Gamma_t = \partial\Omega \setminus \Gamma_u$: $\sigma_{ij}
    n_j = t_i$ on $\Gamma_t$, with $n_j$ the outward unit normal
    and $t_i$ the applied surface traction.
\end{itemize}

The displacement and traction boundary conditions together
constitute the \engterm{boundary-value problem} of elasticity. The
problem is well-posed (existence and uniqueness of the solution)
under mild regularity conditions on the geometry, loads, and
material constants.

\engterm{Plane stress and plane strain}. Two two-dimensional
reductions of the three-dimensional elasticity problem are widely
used.

\emph{Plane stress} applies to thin in-plane-loaded plates, where
the through-thickness stresses are negligible: $\sigma_{zz} =
\sigma_{xz} = \sigma_{yz} = 0$. The displacement is in the $(x, y)$
plane, with the through-thickness strain $\epsilon_{zz}$ determined
by the constitutive equations.

\emph{Plane strain} applies to long prismatic bodies loaded
transversely, where the through-thickness displacement is
constrained: $\epsilon_{zz} = \epsilon_{xz} = \epsilon_{yz} = 0$.
A long pipe under internal pressure (away from the ends), a tunnel
in soil, a long retaining wall: all are plane strain.

The two reductions differ in their effective elastic constants. For
plane stress, $E^* = E$, $\nu^* = \nu$. For plane strain, $E^* =
E/(1 - \nu^2)$, $\nu^* = \nu/(1-\nu)$. The two-dimensional
biharmonic equation $\nabla^4 \phi = 0$ for the Airy stress
function $\phi$ (defined by $\sigma_{xx} = \partial^2\phi/\partial
y^2$, $\sigma_{yy} = \partial^2\phi/\partial x^2$, $\sigma_{xy} =
-\partial^2\phi/\partial x\partial y$) governs both cases; the
difference is in the boundary conditions and in the recovery of
the through-thickness stresses or strains.

The physical difference between the two idealisations decides which one an
analyst is entitled to use, and getting it wrong is one of the quieter ways
a two-dimensional model goes astray. Plane stress is the state of a body
free to breathe out of plane: a thin plate loaded in its plane has nothing
to stop the through-thickness strain, so $\sigma_{zz}$ relaxes to zero and
the material contracts freely in the thickness direction. Plane strain is
the opposite state, of a body prevented from moving out of plane: a slice
deep inside a long dam or a long pipe is held by the identical slices on
either side of it, so the through-thickness strain is zero and a
through-thickness stress $\sigma_{zz} = \nu(\sigma_{xx}+\sigma_{yy})$ builds
up to enforce the constraint. That constraint stress is why plane strain is
the stiffer and more brittle idealisation. The effective Poisson's ratio
$\nu/(1-\nu)$ rises toward one half as $\nu$ grows, so a plane-strain region
behaves as if nearly incompressible, and the triaxial tension it develops at
a crack tip is exactly what makes a thick section more prone to brittle
fracture than a thin one of the same material, the plane-strain fracture
toughness being the lower, conservative material property that design codes
tabulate. The same real component can be in plane stress near its free
surfaces and plane strain in its interior, which is why a fracture that
initiates in the plane-strain core can arrest as it runs out into the
plane-stress skin, leaving the characteristic flat-centre, shear-lip
fracture surface. Choosing plane stress where the body is actually
constrained understates the stiffness and the crack-tip triaxiality;
choosing plane strain where the body is actually free overstates both. The
test is simple: ask whether the out-of-plane direction is free to strain
(plane stress) or held (plane strain), and model accordingly.

\begin{mastery}
\textbf{Saint-Venant's principle and what a boundary condition really
means.} A traction boundary condition specifies the surface stress at
every point, but a real load is rarely known pointwise: a bolt delivers
its force through a contact patch whose pressure distribution no one
measures, a bearing presses through a Hertzian ellipse, a weld carries a
line load whose detail depends on the bead. Saint-Venant's principle
rescues the analyst from needing the detail. It states that two
statically equivalent load distributions, equal in resultant force and
moment, produce stress fields that differ only within a region of order
the size of the loaded patch, and agree everywhere beyond it. A point
load and a pressure patch of the same resultant give the same stress a
few patch-widths away. The principle is the licence behind every
reduced model in this chapter: the beam, the plate, and the membrane
shell all replace a three-dimensional load distribution at an edge or a
joint with its resultant force and moment, and Saint-Venant guarantees
the replacement is accurate away from the edge. It is also the licence
behind the standard finite-element idealisation of a load as a point or
a line, accurate in the field but wrong in the near-region, which is
exactly why the near-region stress at a load application is never read
off the model: it depends on the idealisation, not on the structure.
The principle has a sharp limit. It applies to the decay of a
self-equilibrating load system in a solid body, where the decay length
is the patch size; it does \emph{not} apply to a thin shell, where the
boundary-layer decay length is the much larger $\sqrt{Rh}$ of the shell
section, nor to a slender member in which a self-equilibrating end load
can persist for a length comparable to the cross-section dimension
times the slenderness. An analyst who invokes Saint-Venant to ignore an
edge effect in a thin shell has misapplied it by the ratio of
$\sqrt{Rh}$ to the wall thickness, and that misuse is one more entry in
the catalogue of boundary-condition errors of the failure section.
\end{mastery}

\engterm{The variational principle}. The boundary-value problem of
linear elasticity is equivalent to the minimisation of the
\engterm{total potential energy}
\[
\Pi[\vec{u}] = \int_\Omega \tfrac{1}{2} \sigma_{ij} \epsilon_{ij}
\,dV - \int_\Omega \rho g_i u_i \,dV - \int_{\Gamma_t} t_i u_i \,dA
\]
over all kinematically admissible displacement fields (those that
satisfy the displacement boundary conditions). The displacement
field that minimises $\Pi$ is the unique solution to the
boundary-value problem. The variational formulation is the
foundation of the finite-element method, which discretises the
admissible-displacement space and finds the minimum on the
discretised space.

\subsection{From stress tensor to a yield check}

A solved elasticity problem returns the stress tensor at every point.
The design question is whether that multiaxial stress state will yield
the material, and the answer requires collapsing the six stress
components into a single scalar that can be compared against the
uniaxial yield strength $\sigma_Y$ measured in a tension test. The
two criteria in working use for ductile metals are von Mises and
Tresca.

The \engterm{von Mises criterion} states that yield begins when the
distortion energy reaches the value it has at uniaxial yield. In terms
of the principal stresses $\sigma_1, \sigma_2, \sigma_3$,
\[
\sigma_{\text{vm}} = \sqrt{\tfrac{1}{2}\left[(\sigma_1-\sigma_2)^2 +
(\sigma_2-\sigma_3)^2 + (\sigma_3-\sigma_1)^2\right]},
\]
and yield occurs when $\sigma_{\text{vm}} = \sigma_Y$. In terms of the
full stress tensor the same quantity is
\[
\sigma_{\text{vm}} = \sqrt{\tfrac{1}{2}\left[(\sigma_{xx}-\sigma_{yy})^2
+ (\sigma_{yy}-\sigma_{zz})^2 + (\sigma_{zz}-\sigma_{xx})^2\right]
+ 3\left(\tau_{xy}^2 + \tau_{yz}^2 + \tau_{zx}^2\right)}.
\]
The von Mises stress is the quantity a finite-element post-processor
plots as its default stress contour; the von Mises computation, with a
plane-stress specialisation, is implemented in
\repopath{code/von_mises.py}.

The \engterm{Tresca criterion} states that yield begins when the
maximum shear stress reaches the value at uniaxial yield, $\tau_{\max}
= (\sigma_{\max} - \sigma_{\min})/2 = \sigma_Y/2$. Tresca is the more
conservative of the two; its hexagon is inscribed in the von Mises
ellipse.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-von-mises-envelope}

Figure~\ref{fig:vol03ch09:von-mises-envelope} plots both envelopes in
the plane of the two in-plane principal stresses. The largest gap
between them, a factor of $2/\sqrt{3} \approx 1.155$, occurs in pure
shear; for most biaxial states the two criteria agree to within ten
percent. An engineer reading an FEM von Mises plot is reading the
distance of the stress state from the boundary of this ellipse.

\section{Anisotropy preview}\pathtag{mastery}

The isotropic constitutive law $\sigma_{ij} = \lambda \delta_{ij}
\epsilon_{kk} + 2\mu \epsilon_{ij}$ assumes that the material's
elastic response is independent of orientation. The assumption
fails for many engineering materials: wood (orthotropic), composite
laminates (orthotropic at the lamina, more general at the
laminate), single crystals (cubic, hexagonal, or lower symmetry),
rolled steel (mild orthotropy from rolling texture), and biological
tissues (anisotropy that varies with loading history).

The general linear-elastic constitutive law for an anisotropic
material has the form
\[
\sigma_{ij} = C_{ijkl} \epsilon_{kl},
\]
where $C_{ijkl}$ is the fourth-rank elasticity tensor with
$3^4 = 81$ components. The symmetries $\sigma_{ij} = \sigma_{ji}$,
$\epsilon_{kl} = \epsilon_{lk}$, and (from the existence of a
strain-energy potential) $C_{ijkl} = C_{klij}$ reduce the number
of independent components to $21$ in the most general triclinic
case. Material symmetries further reduce the count: orthotropic
materials have $9$ independent constants; tetragonal have $6$;
hexagonal (or transversely isotropic) have $5$; cubic have $3$;
isotropic have $2$.

\engterm{Wood} is approximately orthotropic with three principal
directions: longitudinal (along the grain), radial, and tangential
to the growth rings. The longitudinal Young's modulus exceeds the
radial and tangential moduli by factors of $20$ or more; the shear
moduli are smaller again. The structural engineer who designs in
wood designs in the orthotropic frame and respects the
grain-direction-dependent allowable stresses (NDS for the United
States, Eurocode 5 for the European Union).

\engterm{Composite laminates} are constructed from layers of
unidirectional fibre-reinforced material stacked at chosen
orientations. A typical carbon-fibre-epoxy lamina has
longitudinal (fibre-direction) Young's modulus of $130$ to $200\,
\text{GPa}$, transverse Young's modulus of $7$ to $10\,\text{GPa}$,
in-plane shear modulus of $4$ to $7\,\text{GPa}$, and major
Poisson's ratio of $0.25$ to $0.35$. The laminate's effective
elastic properties depend on the stacking sequence, layer
thicknesses, and ply orientations; the working tool is classical
laminate plate theory, which assembles the laminate stiffness
matrix from the lamina stiffness matrices and the ply-orientation
rotations.

Classical laminate theory collects the laminate's response into a single
$6\times 6$ relation, written in three blocks: an extensional stiffness
$\mathbf{A}$ relating in-plane forces to in-plane strains, a bending
stiffness $\mathbf{D}$ relating moments to curvatures, and a coupling
stiffness $\mathbf{B}$ that ties the two together. The coupling block is
where laminate design departs from the intuition an isotropic plate builds.
A laminate whose stacking is not symmetric about its midplane has a nonzero
$\mathbf{B}$, and a nonzero $\mathbf{B}$ means that pulling on the laminate
bends it and heating it warps it, because the plies above the midplane
respond differently from those below. An unsymmetric laminate cured flat
comes out of the oven curved, from the same thermal mismatch that curls a
bimetallic strip. A laminate whose off-axis plies are not balanced (equal
numbers at $+\theta$ and $-\theta$) twists when stretched, the
extension-shear coupling that has no isotropic analogue. The design
discipline is therefore to choose the stacking sequence for the couplings it
turns on or off: a symmetric balanced laminate ($\mathbf{B}=0$, no
extension-shear coupling) behaves like an orthotropic plate and is the safe
default, while an engineered coupling is exploited deliberately, as in the
aeroelastically tailored wing skin whose bend-twist coupling washes the wing
out of high-angle-of-attack loads. The lesson for the finite-element analyst
is that a composite plate model carries a material description far richer
than two constants, and getting the layup wrong changes not only the
stiffness magnitude but the very modes the plate couples, in a way no
isotropic model can approximate.

The design discipline for anisotropic materials starts with the
correct material model. An isotropic FEM analysis of a composite
laminate, or a wooden beam loaded across the grain analysed as
isotropic, gives wrong stresses by factors that can exceed two and
unsafe predictions of failure. Composite design is the subject of
Volume V chapters 11 and 12; the present chapter notes only that
the FEM machinery extends to anisotropic materials by replacing
the isotropic elasticity tensor with the anisotropic tensor,
nothing more.

The most consequential anisotropic plate in everyday structures is the
\engterm{sandwich panel}, two thin stiff facesheets bonded to a thick
light core (figure~\ref{fig:vol03ch09:sandwich-panel}). The construction
exploits the thickness-cubed scaling of flexural rigidity in reverse:
rather than thicken a solid plate, it moves the load-carrying face
material as far from the midsurface as the core allows. For facesheets of
thickness $t_f$ separated by a centroid spacing $d$, the bending
stiffness per unit width is dominated by the parallel-axis term
$D \approx E_f t_f d^2/2$, which grows as the square of the separation,
while the panel's mass grows only linearly with the core depth. A
sandwich with a separation ten times the facesheet thickness is roughly
two orders of magnitude stiffer in bending than a single facesheet of the
same mass per unit area. The core's job is to carry the transverse shear
and to hold the faces apart against it, so a low-density honeycomb or
foam suffices, and the panel is strongly orthotropic: stiff in bending,
weak in transverse shear. The price is a set of failure modes the solid
plate never has. The compression-side face can wrinkle into short
wavelengths that the global bending stress does not flag, the core can
shear or crush, and the bond can debond, each governed by the core and
adhesive properties rather than by the face material. A sandwich panel
analysed with the solid-plate bending check alone passes on stiffness and
fails in service on a mode the check never examined, the same lesson the
stiffened panel of worked example 7 taught in its discrete form.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-sandwich-panel}

\begin{mastery}
\textbf{The elasticity tensor in Voigt notation.} The fourth-rank
tensor $C_{ijkl}$ with its $21$ independent constants is awkward to
carry around, so engineering practice flattens the symmetric stress
and strain tensors into six-vectors and writes the constitutive law as
a $6\times 6$ symmetric matrix. The \engterm{Voigt convention} maps the
index pairs $(11,22,33,23,31,12) \to (1,2,3,4,5,6)$, so that
\[
\begin{pmatrix} \sigma_1 \\ \sigma_2 \\ \sigma_3 \\ \sigma_4 \\
\sigma_5 \\ \sigma_6 \end{pmatrix}
= \begin{pmatrix}
C_{11} & C_{12} & C_{13} & C_{14} & C_{15} & C_{16} \\
       & C_{22} & C_{23} & C_{24} & C_{25} & C_{26} \\
       &        & C_{33} & C_{34} & C_{35} & C_{36} \\
       &        &        & C_{44} & C_{45} & C_{46} \\
       &        &        &        & C_{55} & C_{56} \\
\text{sym} & & & & & C_{66}
\end{pmatrix}
\begin{pmatrix} \epsilon_1 \\ \epsilon_2 \\ \epsilon_3 \\
\epsilon_4 \\ \epsilon_5 \\ \epsilon_6 \end{pmatrix},
\]
where the shear strains in the strain six-vector are the engineering
shears $\gamma_{ij} = 2\epsilon_{ij}$. The matrix has $21$ independent
entries in the triclinic case. For an orthotropic material aligned with
the coordinate axes, the off-diagonal shear couplings $C_{14}$ through
$C_{56}$ vanish and the upper-left $3\times 3$ block plus three
diagonal shear terms leave $9$ constants. For isotropy the matrix
collapses to two: $C_{11} = C_{22} = C_{33} = \lambda + 2\mu$,
$C_{12} = C_{13} = C_{23} = \lambda$, $C_{44} = C_{55} = C_{66} = \mu$.
Every commercial finite-element code stores the material as this
$6\times 6$ matrix; choosing the right one for the material at hand is
the single most consequential material-model decision, because an
isotropic matrix applied to a composite laminate is wrong by the ratio
of the longitudinal to transverse modulus, a factor that often exceeds
ten.
\end{mastery}

\section{Finite-element method: introduction}\pathtag{core}

The \engterm{finite-element method} (FEM) is a numerical method
for solving partial differential equations on arbitrary geometries.
For linear elasticity, the FEM discretises the domain into
elements, approximates the displacement field within each element
as a low-order polynomial in the element's local coordinates,
assembles the global stiffness matrix from the elements, and
solves the resulting linear system for the nodal displacements.
The method was developed in the 1950s and 1960s for aerospace
structural analysis (Turner, Clough, Martin, Topp, 1956) and has
become the dominant analytical tool for structural mechanics
across every engineering discipline. The standard engineering
reference is Zienkiewicz, Taylor, and Zhu
\cite[chs.~1--3]{text:zienkiewicz-fem}.

The method's working steps, in the order an engineer performs
them:

\begin{enumerate}
  \item \engterm{Discretise the geometry.} Divide the body into
    elements: triangles or quadrilaterals in two dimensions;
    tetrahedra, hexahedra, prisms, or pyramids in three dimensions.
    The collection of elements is the \engterm{mesh}; the points at
    which elements connect are the \engterm{nodes}.
  \item \engterm{Choose the element type.} The element type fixes
    the polynomial order of the displacement field within the
    element. Linear (first-order) elements have nodes only at
    corners; quadratic (second-order) elements have additional
    nodes at edge midpoints. Higher-order elements give more
    accuracy per element but more computation per element.
  \item \engterm{Assemble the global stiffness matrix.} Each
    element contributes a small stiffness matrix $K_e$ (typically
    $6 \times 6$ to $30 \times 30$ depending on element type and
    dimension) that relates the element's nodal forces to its
    nodal displacements. The global stiffness matrix $K$ is
    assembled by summing the element contributions at the shared
    nodes; the result is a large sparse symmetric positive
    semidefinite matrix.
  \item \engterm{Impose boundary conditions.} Displacement
    boundary conditions are imposed by removing the corresponding
    rows and columns of $K$ and the corresponding entries of the
    force vector $\vec{F}$. Traction boundary conditions become
    nodal forces by integrating the traction over the element
    boundary.
  \item \engterm{Solve the linear system} $K \vec{u} = \vec{F}$
    for the nodal displacements $\vec{u}$. The solver is either
    direct (sparse Cholesky factorisation) for problems up to
    perhaps $10^7$ degrees of freedom, or iterative (conjugate
    gradient with preconditioning) for larger problems.
  \item \engterm{Recover element stresses and strains} from the
    nodal displacements by differentiating the element's
    interpolated displacement field. The stresses are typically
    most accurate at the Gauss points (interior integration
    points) of the element and less accurate at the nodes and
    edges.
\end{enumerate}

The method is general: the same six steps work for plates, shells,
three-dimensional solids, plane stress, plane strain, axisymmetric
problems, linear and nonlinear materials, static and dynamic loads.
Every commercial finite-element package (Abaqus, ANSYS, NASTRAN,
LS-DYNA, COMSOL, the open-source Code\_Aster and CalculiX) implements
the same machinery with different element libraries and different
solvers. A teaching-grade implementation of the six steps for
plane-stress problems with linear triangular elements, short enough to
read in one sitting, is in \repopath{code/cst_fem.py}; running it on
its own patch test reproduces the analytical uniaxial strain
$\epsilon = \sigma/E$ and makes the assembly-and-solve loop concrete
before any commercial package hides it.

\engterm{Convergence}. The FEM solution depends on the mesh: refining
the mesh changes the answer, and a well-posed problem produces a
sequence of FEM solutions that converges to the exact solution as
the mesh is refined. The standard refinement strategies:

\begin{itemize}[leftmargin=*]
  \item \engterm{h-refinement}: keep the element order fixed, reduce
    the element size. The error decreases as $h^p$, where $p$ is
    the polynomial order of the element and $h$ is the
    characteristic element size.
  \item \engterm{p-refinement}: keep the mesh fixed, increase the
    element order. The error decreases exponentially in $p$ for
    smooth solutions.
  \item \engterm{hp-refinement}: refine the mesh near features
    (corners, point loads, material discontinuities) and increase
    the element order in smooth regions. The combination gives the
    best convergence rate but requires more sophisticated mesh
    management.
\end{itemize}

\engterm{Mesh-convergence study}. A working FEM analysis is not a
single run on a single mesh; it is a sequence of runs on progressively
refined meshes, with the engineer verifying that the quantity of
interest (peak stress, maximum deflection, natural frequency) is
converging to a stable value as the mesh is refined. A typical study
runs the analysis on three or four meshes with element-size ratios
of $2:1$, plots the quantity of interest against the inverse element
size, and extrapolates to zero element size by Richardson
extrapolation. A result that has not been checked against mesh
convergence is a result that depends on the mesh, and is therefore
not a result.

\begin{mastery}
\textbf{Where the element stiffness matrix comes from.} The element
stiffness matrix $K_e$ is not a black box; it is the discretised form
of the variational principle. Within an element, the displacement field
is interpolated from the nodal displacements $\vec{d}_e$ by the
\engterm{shape functions} $N_a(\vec{x})$,
\[
\vec{u}(\vec{x}) = \sum_a N_a(\vec{x})\,\vec{d}_{e,a}
= \mathbf{N}(\vec{x})\,\vec{d}_e .
\]
Differentiating gives the strain in terms of the nodal displacements
through the \engterm{strain-displacement matrix} $\mathbf{B} =
\partial \mathbf{N}$, so $\boldsymbol{\epsilon} = \mathbf{B}\,
\vec{d}_e$. Substituting into the strain energy
$\tfrac{1}{2}\int_{\Omega_e} \boldsymbol{\epsilon}^{\mathsf T}
\mathbf{D}\,\boldsymbol{\epsilon}\,dV$ and minimising over $\vec{d}_e$
yields
\[
K_e = \int_{\Omega_e} \mathbf{B}^{\mathsf T}\,\mathbf{D}\,\mathbf{B}
\,dV ,
\]
with $\mathbf{D}$ the material matrix above. For a constant-strain
triangle, $\mathbf{B}$ is constant over the element and the integral is
just $\mathbf{B}^{\mathsf T}\mathbf{D}\,\mathbf{B}$ times the element
area and thickness, which is exactly the arithmetic implemented in
\repopath{code/cst_fem.py}. For higher-order elements $\mathbf{B}$
varies over the element and the integral is evaluated numerically at
Gauss points, which is why element stresses are most accurate there.
Assembly is the statement that the displacement is continuous across
element boundaries: the global $K$ is the sum of element $K_e$ with
shared nodes overlapped. Every commercial code does precisely this; the
sophistication is in the element library ($\mathbf{B}$ formulations
that avoid locking) and the sparse solver, not in the recipe.
\end{mastery}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-mesh-convergence}

Figure~\ref{fig:vol03ch09:mesh-convergence} shows the study for the
plate-with-a-hole problem that recurs as the chapter project: the peak
stress climbs from a coarse-mesh under-prediction toward the
analytical value as the mesh refines. The coarse mesh is wrong by
about fifteen percent and on the unsafe side. The Richardson
extrapolation that recovers the mesh-independent value from three runs
is implemented in \repopath{code/richardson_convergence.py}; the
underlying mesh-by-mesh data are in
\repopath{data/mesh_convergence.csv}. The structure of the curve is
worth committing to memory: stress quantities at concentrations
typically converge from below, so a single coarse run does not merely
have an unknown error, it has an error of known sign that flatters the
design.

\begin{archetype}[Stability]
The archetype of plate, shell, and three-dimensional elasticity is
\emph{stability}: the load capacity of the structure is set not by
material strength but by the geometric stability of the loaded
configuration. A thin spherical shell buckles at $20\,\%$ of its
linear-theory critical load because a small dimple destabilises
the membrane equilibrium. A thin-walled tube under bending
ovalises and triggers the Brazier failure at a moment below the
linear prediction. A plate under in-plane compression buckles
out-of-plane at a load far below the material's yield. The
discipline of stability-dominated design is to compute not only
the static stress state but also the bifurcation point of the
loaded equilibrium, and to design with a stability margin large
enough to absorb the imperfections that real structures carry.
\end{archetype}

\section{Mesh quality and the engineer's responsibility}\pathtag{core}

A finite-element solution is only as good as the mesh on which it
is computed. A poor mesh produces wrong answers, and the wrongness
is not signalled by the solver: the analysis runs, the colours on
the screen look reassuring, and the numbers are wrong. The
engineer's responsibility is to verify the mesh before trusting
the result.

The diagnostics of mesh quality, in order of importance:

\engterm{Element aspect ratio.} The ratio of the element's longest
to its shortest edge. For linear tetrahedra and triangles, aspect
ratios above $\sim 5$ produce unreliable stress recovery;
ratios above $\sim 20$ produce wrong results. The mesh generator
must control aspect ratio everywhere, including near boundaries
and geometric features where automatic mesh refinement may
inadvertently produce thin slivers.

\engterm{Element skewness.} The deviation of the element's
internal angles from their ideal values ($60^{\circ}$ for
equilateral triangles, $90^{\circ}$ for quadrilaterals, the
corresponding values for three-dimensional elements). Skewed
elements have poorly conditioned stiffness matrices and produce
inaccurate stresses. The working rule is to keep all interior
angles within $\pm 30^{\circ}$ of the ideal.

\engterm{Element warpage} (for quadrilateral and hexahedral
elements). A quadrilateral element whose four corner nodes are
not coplanar is warped; the warpage produces an additional
mode in the element's stiffness matrix that distorts the
displacement field. The working rule is to keep the warpage
below $5\,\%$ of the element's characteristic size.

\engterm{Mesh refinement near stress concentrations.} A hole, a
sharp corner, a material discontinuity, a point load, or any
location of high stress gradient requires fine mesh refinement
locally. A coarse mesh at a stress concentration under-predicts
the peak stress by factors that can exceed two; an FEM analysis
that misses a stress concentration is an FEM analysis that
predicts a non-failure for a member that will fail.

\engterm{Convergence of the quantity of interest.} The peak stress
at a stress concentration may converge slowly with mesh
refinement, or it may not converge at all if the geometry has a
re-entrant corner (a singular stress field). The engineer must
verify convergence on the quantity of interest, not on any
quantity. A converged maximum-deflection result is not a
converged maximum-stress result; the two quantities have different
convergence rates and may require different meshes.

\engterm{Material-discontinuity meshing.} Elements should not span
material discontinuities. A composite laminate is meshed with
separate element layers for each ply; a steel-concrete composite
beam is meshed with separate elements for the steel and the
concrete, connected at the interface by either continuous
displacement or a contact-mechanics interface.

\engterm{Element selection.} For thin structures (plates, shells)
the engineer uses shell elements rather than solid elements,
because solid elements with high aspect ratio in the thickness
direction are prone to \engterm{shear locking}, a numerical
artifact in which the element's bending stiffness is overestimated
by a factor that can exceed ten. The shell element formulation
handles the thin-structure kinematics correctly. Similarly, for
nearly incompressible materials ($\nu \to 0.5$, as in rubber or
nearly-saturated soil), the engineer uses elements formulated with
mixed displacement-pressure interpolation to avoid
\engterm{volumetric locking}.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-shear-locking}

Figure~\ref{fig:vol03ch09:shear-locking} shows the mechanism of shear
locking, the most common way a beginner's solid-element model of a
thin part comes back far too stiff. A single linear element through the
thickness has straight edges and cannot bend; it manufactures a
parasitic transverse shear strain that soaks up energy and reports a
deflection that can be a half or a third of the truth. The diagnosis
is to compare against the analytical Kirchhoff prediction, the cure is
shell elements or a stack of quadratic elements.

\begin{mastery}
\textbf{Reduced integration, hourglassing, and the cure that has its own
disease.} The element stiffness matrix $K_e = \int_{\Omega_e}
\mathbf{B}^{\mathsf T}\mathbf{D}\,\mathbf{B}\,dV$ is evaluated numerically at
a set of Gauss integration points. Full integration uses enough points to
compute the polynomial integrand exactly, and it is full integration that
produces shear and volumetric locking: the extra integration points sample
the parasitic strain modes and let them absorb energy, so the element comes
back too stiff. The standard cure is \engterm{reduced integration}, using
fewer Gauss points than exact integration would require. With one point at
the centre of a linear quadrilateral, the parasitic shear strain, which
happens to vanish at the centre, is never sampled, so the element bends
freely and the locking is gone. Reduced integration also runs faster,
because it evaluates $\mathbf{B}^{\mathsf T}\mathbf{D}\,\mathbf{B}$ at fewer
points. The cure carries its own disease. A single integration point cannot
detect deformation modes that leave that one point unstrained, and a linear
quadrilateral has exactly such modes: the element can distort into an
hourglass shape (the two diagonals lengthening and shortening) with zero
strain at its centre, so the stiffness matrix assigns that mode no energy.
These zero-energy \engterm{hourglass modes} are spurious mechanisms, and
under load they can grow without restraint, filling the mesh with a
jagged, alternating pattern that looks like a numerical instability because
it is one. The fix is \engterm{hourglass control}, a small artificial
stiffness added to resist the zero-energy modes without restoring the
locking that reduced integration removed. Every commercial code offers
reduced-integration elements with hourglass control as its workhorse, and
the analyst's responsibility is to watch for the tell-tale hourglass pattern
in the deformed shape and to confirm that the artificial hourglass energy
stays a small fraction of the total strain energy, the code's own diagnostic
that the control is doing its job and not quietly stiffening the answer. The
full-integration locking of figure~\ref{fig:vol03ch09:shear-locking} and the
reduced-integration hourglassing are the two horns of the same dilemma, and
the modern element formulations (assumed strain, enhanced strain, mixed
interpolation) exist to escape both at once.
\end{mastery}

\engterm{The mesh-generation workflow.} A working FEM analyst
proceeds:

\begin{enumerate}
  \item Generate a first coarse mesh on the entire domain, with
    element size comparable to the smallest geometric feature.
  \item Refine locally near stress concentrations, geometric
    discontinuities, and material interfaces.
  \item Check element quality (aspect ratio, skewness, warpage)
    against the analysis software's quality metrics. Repair
    poor-quality elements either manually or by remeshing.
  \item Run the analysis on the first mesh; record the quantity
    of interest.
  \item Refine the mesh by a factor of two; rerun; record the
    quantity of interest.
  \item Compare the two runs. If they agree to within the
    required tolerance, the analysis is converged; if not,
    refine again. Continue until convergence is confirmed.
\end{enumerate}

A finite-element analysis whose mesh-convergence study is not in
the report is not a finite-element analysis whose results can be
trusted.

\section{Verification, validation, uncertainty quantification}\pathtag{standard}

A finite-element analysis produces numerical answers. Whether
those numerical answers correspond to the physical system the
engineer wishes to predict is a separate question, addressed by
the disciplines of \engterm{verification}, \engterm{validation},
and \engterm{uncertainty quantification} (VVUQ).

\engterm{Verification} asks: does the simulation solve the
mathematical model correctly? It is the discipline of confirming
that the software, the mesh, the solver, and the numerical
parameters together produce the solution of the boundary-value
problem the engineer intended to solve. Verification is internal
to the simulation; it does not require any comparison with physical
reality. The standard verification tools:

\begin{itemize}[leftmargin=*]
  \item \engterm{Code verification}. Run the software on test
    problems with known analytical solutions and confirm that the
    numerical solution converges to the analytical solution at
    the expected rate. Commercial codes are verified by their
    vendors against benchmark suites; the engineer should be
    aware of the verified element types and physical models in
    use.
  \item \engterm{Calculation verification}. Run a mesh-convergence
    study (section 9.6) on the actual problem. Confirm that the
    quantity of interest converges to a stable value as the mesh
    is refined.
  \item \engterm{Order-of-accuracy verification}. Confirm that the
    convergence rate matches the theoretical rate for the element
    type. A linear element should give $O(h^2)$ convergence in
    energy norm; a quadratic element should give $O(h^4)$. A
    convergence rate below the theoretical rate signals a problem
    (mesh quality, material model, boundary condition
    implementation).
\end{itemize}

\engterm{Validation} asks: does the mathematical model describe
the physical system? It is the discipline of comparing the
simulation result to experimental measurement on the physical
system. Validation is external to the simulation; it requires
physical data. The standard validation methods:

\begin{itemize}[leftmargin=*]
  \item \engterm{Validation against a controlled test}. A simple
    component is tested under controlled load; the test measures
    deflection, strain at instrumented points, and failure load.
    The simulation predicts the same quantities; the agreement
    (or disagreement) is recorded.
  \item \engterm{Validation hierarchy}. A complex system is
    validated by validating subcomponents at increasing levels of
    complexity: material coupons, structural elements, sub-
    assemblies, and finally the full system. The hierarchy gives
    confidence that the simulation captures the relevant physics
    at every scale.
  \item \engterm{Validation under representative loading}. The
    validation test loads the component in conditions that
    represent the intended service: temperature range, strain
    rate, environment. A validation under static room-temperature
    conditions does not validate the simulation for use at high
    temperature or under impact loading.
\end{itemize}

\engterm{Uncertainty quantification} asks: how much do the inputs
vary, and how much does the output vary in response? It is the
discipline of propagating input uncertainties (material constants,
geometric tolerances, load magnitudes) through the simulation to
produce a distribution of outputs. The working tools:

\begin{itemize}[leftmargin=*]
  \item \engterm{Sensitivity analysis}. Compute the partial
    derivatives of the quantity of interest with respect to each
    input. The largest derivatives identify the inputs that most
    affect the output.
  \item \engterm{Monte Carlo sampling}. Sample the input
    distributions, run the simulation for each sample, and
    accumulate the output distribution. Monte Carlo gives the
    output distribution exactly in the limit of many samples but
    is expensive for slow simulations.
  \item \engterm{Polynomial chaos expansion}. Represent the
    output as a polynomial expansion in the input random variables
    and compute the expansion coefficients from a small number of
    simulations. The method is far more efficient than Monte Carlo
    for smooth input-output relationships.
\end{itemize}

The choice among the three tools is a budget decision, and the budget is
counted in simulation runs. A single finite-element analysis of a real
structure can take minutes to hours, so the number of runs the uncertainty
study can afford is often small, and the three methods trade accuracy for
runs in different ways. Sensitivity analysis is the cheapest, needing one
run per input to estimate the local derivatives by finite differences, and
it answers the first question worth asking: which inputs matter. Its
limitation is that it is local, valid near the nominal point, and it can
miss a strong interaction that only appears when two inputs move together.
Monte Carlo is the most general and the most expensive, converging on the
output distribution as the reciprocal square root of the number of runs, so
halving the error costs four times the runs, which is punishing for a slow
simulation. Polynomial chaos threads between the two: for a smooth response
in a handful of random inputs it captures the whole output distribution from
tens of runs rather than thousands, but it degrades as the number of
uncertain inputs grows (the curse of dimensionality) and it can mislead when
the response is not smooth, as at a buckling bifurcation or a contact that
opens and closes. A working uncertainty study usually chains them: a
sensitivity analysis first to rank the inputs and discard the ones that do
not matter, then a polynomial chaos or a focused Monte Carlo on the two or
three inputs that survived. The output that discipline produces is the one an
engineering claim needs: not a single deflection or peak stress, but a
nominal value with a quantified spread traceable to the input tolerances
that drive it, which is what lets a designer state a probability of
exceeding an allowable rather than a bare number that hides its own
uncertainty.

The VVUQ discipline is the discipline that turns a finite-element
result into a quantitative engineering claim. A simulation without
verification produces a number; a simulation with verification but
without validation produces a number that solves the wrong problem;
a simulation with verification and validation but without
uncertainty quantification produces a number whose error bars are
unknown. The full VVUQ pipeline produces a number with error bars,
calibrated against physical reality. The full pipeline is what
the regulatory authority will require for any safety-critical
structural certification.

\begin{estimation}
\textbf{Estimate the central deflection of a $4 \,\text{m} \times 4
\,\text{m}$ square aluminium plate, $5 \,\si{\milli\meter}$ thick,
simply supported on all four edges, carrying a uniform pressure of
$1\,\text{kPa}$.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
For a simply supported square plate of side $a$ under uniform
pressure $q_0$, the centre deflection from the Navier series is
$w_{\max} \approx 0.00406 q_0 a^4/D$, where $D = E h^3/[12(1-\nu^2)]$.

For aluminium ($E \approx 70\,\text{GPa}$, $\nu \approx 0.33$),
\[
D = \frac{70 \times 10^9 \times (5 \times 10^{-3})^3}{12 \times
(1 - 0.33^2)} \approx \frac{70 \times 10^9 \times 1.25 \times 10^{-7}}{
10.7} \approx 818\,\text{N}\cdot\text{m}.
\]
With $q_0 = 10^3\,\text{Pa}$, $a = 4\,\text{m}$, $a^4 = 256\,
\text{m}^4$,
\[
w_{\max} \approx 0.00406 \times \frac{10^3 \times 256}{818}
\approx 1.27\,\text{m}.
\]
That answer is absurd: a deflection of more than a metre at the
centre of a plate $5\,\text{mm}$ thick spanning $4\,\text{m}$ violates
the small-deflection assumption of Kirchhoff theory by orders of
magnitude. The estimate flags the design as a non-starter: the
plate is far too thin for its span.

The postmortem: the estimate has done its job by showing that
linear plate theory does not apply. The plate is in the
membrane-action regime, in which the deflected midsurface stretches
and carries the load by tensile in-plane stresses rather than by
bending. The membrane analysis of a thin square plate gives a
deflection of order $w_{\max} \sim a (q_0 a/(Eh))^{1/3}$. With the
above numbers, $w_{\max} \sim 4 \times (10^3 \times 4/(70 \times
10^9 \times 5 \times 10^{-3}))^{1/3} \approx 4 \times (1.14 \times
10^{-5})^{1/3} \approx 4 \times 0.022 \approx 0.09\,\text{m}$. The
plate deflects $\sim 90\,\si{\milli\meter}$ at the centre,
which is plausible for a $5\,\si{\milli\meter}$ membrane spanning
$4\,\text{m}$ under $1\,\text{kPa}$, but is still likely beyond the
serviceability limit for any building application. The honest
design specifies a thicker plate, stiffeners, or both.

The deeper lesson: a thin plate's linear bending analysis is valid
only when the deflection is less than the plate's thickness. A
deflection comparable to or larger than the thickness moves the
plate into the membrane regime, and a different model is required.
A finite-element analysis run in linear mode on this geometry would
give the same absurd answer as the formula; an analyst who trusted
the result without checking against the small-deflection criterion
would specify a plate that fails in service. The deflection-to-
thickness ratio is the load-bearing diagnostic.
\end{estimation}

\begin{mastery}
\textbf{Large deflection: the von K\'arm\'an coupling.} The estimation
above failed because the linear plate equation $D\nabla^4 w = q$ assumes
the midsurface neither stretches nor carries in-plane force, an
assumption valid only while the deflection stays below the plate
thickness. Once the deflection grows comparable to the thickness, the
deflected midsurface must stretch to take its new shape, and the
stretching generates in-plane membrane forces that carry part of the
load directly, in parallel with the bending. The governing equations
become the coupled pair due to von K\'arm\'an,
\[
D\nabla^4 w = q + h\!\left(
\frac{\partial^2\phi}{\partial y^2}\frac{\partial^2 w}{\partial x^2}
- 2\frac{\partial^2\phi}{\partial x\partial y}
\frac{\partial^2 w}{\partial x\partial y}
+ \frac{\partial^2\phi}{\partial x^2}\frac{\partial^2 w}{\partial y^2}
\right),
\qquad
\nabla^4\phi = E\!\left[
\left(\frac{\partial^2 w}{\partial x\partial y}\right)^2
- \frac{\partial^2 w}{\partial x^2}\frac{\partial^2 w}{\partial y^2}
\right],
\]
where $\phi$ is the in-plane (Airy) stress function carrying the membrane
forces and the right-hand sides couple the out-of-plane deflection $w$ to
the in-plane stretching. The first equation is the plate bending equation
with a membrane-restoring term added; the second says that any
double-curvature in $w$ (a shape that cannot be flattened without
stretching) sources membrane stress. The pair is nonlinear, because the
products of second derivatives of $w$ appear, and that nonlinearity is
the reason the linear superposition that works below the thickness scale
breaks above it. Two consequences run through plate design. The plate is
stiffer than linear theory predicts once the deflection grows, because
the membrane action shares the load, which is why the absurd
metre-and-more linear estimate of the worked estimation collapsed to a
plausible $90\,\si{\milli\meter}$ once membrane action was admitted.
And the in-plane membrane force the deflection induces is exactly the
force that can drive post-buckling and oil-canning in a compressed or
restrained panel, the bridge between this chapter's bending theory and
its stability theory. A finite-element analysis run in linear mode on a
plate that deflects past its thickness gives the same wrong answer the
formula does; the cure is to switch on geometric nonlinearity, which
turns on these very coupling terms inside the element formulation.
\end{mastery}

\section{Worked examples}\pathtag{standard}

The formulas of the preceding sections become engineering only when
they are pushed through to numbers with units, checked for plausibility,
and turned into a decision. The examples below work the canonical cases
end to end: a plate deflection, a pressure-vessel wall sizing, a
multiaxial yield check, a stress concentration, and a shell-buckling
margin. Each is the kind of calculation that a working engineer
performs on the back of a free-body diagram before any finite-element
package is opened.

\subsection*{Worked example 1: deflection of a simply supported plate}

\textbf{Problem.} A simply supported square steel plate, side
$a = 1\,\si{\meter}$, thickness $h = 10\,\si{\milli\meter}$, $E =
200\,\text{GPa}$, $\nu = 0.3$, carries a uniform pressure $q_0 =
5\,\text{kPa}$. Find the centre deflection and the maximum surface
stress, and confirm that linear plate theory applies.

\textbf{Solution.} The flexural rigidity is
\[
D = \frac{E h^3}{12(1-\nu^2)} = \frac{200\times 10^9 \times (0.01)^3}
{12(1-0.09)} = \frac{200\times 10^9 \times 10^{-6}}{10.92}
= 1.832\times 10^4\,\text{N}\cdot\text{m}.
\]
For a simply supported square plate the centre deflection is
$w_{\max} = 0.00406\, q_0 a^4/D$, so
\[
w_{\max} = 0.00406 \times \frac{5\times 10^3 \times 1^4}
{1.832\times 10^4} = 1.11\times 10^{-3}\,\text{m} = 1.11\,\si{\milli\meter}.
\]
The maximum bending moment per unit length at the centre is
$M_{\max} \approx 0.0479\, q_0 a^2 = 0.0479\times 5000\times 1 =
240\,\text{N}\cdot\text{m}/\text{m}$. The surface stress follows from
$\sigma_{\max} = 6M_{\max}/h^2$:
\[
\sigma_{\max} = \frac{6\times 240}{(0.01)^2} = 1.44\times 10^7\,
\text{Pa} = 14.4\,\text{MPa},
\]
comfortably below the $250\,\text{MPa}$ yield of structural steel,
giving a yield margin of about $17$. The deflection-to-thickness ratio
is $1.11/10 = 0.11$, well under unity, so the small-deflection
Kirchhoff theory is valid and the plate is in the bending regime, not
the membrane regime. The Navier-series cross-check, summed in
\repopath{code/navier_plate.py}, reproduces the $0.00406$ coefficient
to four figures.

\subsection*{Worked example 2: sizing a cylindrical pressure vessel}

\textbf{Problem.} A cylindrical steel pressure vessel of inner radius
$R = 0.4\,\si{\meter}$ must hold an internal gauge pressure
$p = 4\,\text{MPa}$ with a safety factor of $3$ against yield
($\sigma_Y = 250\,\text{MPa}$). Size the cylindrical wall, then size a
hemispherical end cap, and compare.

\textbf{Solution.} The governing stress in the cylinder is the hoop
stress $\sigma_\theta = pR/h$. The allowable stress is $\sigma_{\text{
allow}} = \sigma_Y/3 = 83.3\,\text{MPa}$. Setting $\sigma_\theta =
\sigma_{\text{allow}}$ and solving for the wall thickness,
\[
h_{\text{cyl}} = \frac{pR}{\sigma_{\text{allow}}} = \frac{4\times 10^6
\times 0.4}{83.3\times 10^6} = 0.0192\,\text{m} = 19.2\,\si{\milli\meter}.
\]
The hemispherical cap carries $\sigma = pR/2h$, half the cylinder hoop
for the same wall, so
\[
h_{\text{cap}} = \frac{pR}{2\sigma_{\text{allow}}} = \frac{4\times 10^6
\times 0.4}{2\times 83.3\times 10^6} = 9.6\,\si{\milli\meter},
\]
exactly half the cylinder wall. The thinner cap is the structural
reason hemispherical and ellipsoidal heads are preferred over flat
ones, and the reason the cylinder-to-head junction, where the two wall
thicknesses and the two stress states meet, is the location of a
bending boundary layer that the membrane theory alone does not
capture; a real design adds a knuckle radius and a local thickness
transition there. The sizing arithmetic is in
\repopath{code/pressure_vessel.py}.

\subsection*{Worked example 3: a multiaxial yield check}

\textbf{Problem.} A point in the wall of the cylinder above carries
hoop stress $\sigma_\theta = 125\,\text{MPa}$, axial stress $\sigma_z =
62.5\,\text{MPa}$, and a small radial compression averaging
$\sigma_r \approx -2\,\text{MPa}$ across the wall. Compute the von
Mises stress and the safety factor against the $250\,\text{MPa}$
yield. Compare with the Tresca prediction.

\textbf{Solution.} The three values are principal stresses (the wall
state is biaxial with a small radial term), so
\[
\sigma_{\text{vm}} = \sqrt{\tfrac{1}{2}\left[(125-62.5)^2 +
(62.5-(-2))^2 + ((-2)-125)^2\right]}\ \text{MPa}.
\]
The bracket is $\tfrac{1}{2}[62.5^2 + 64.5^2 + 127^2] =
\tfrac{1}{2}[3906 + 4160 + 16129] = 12098$, so $\sigma_{\text{vm}} =
\sqrt{12098} = 110\,\text{MPa}$. The safety factor against yield is
$250/110 = 2.27$. The Tresca prediction uses the maximum shear:
$\tau_{\max} = (\sigma_{\max} - \sigma_{\min})/2 = (125-(-2))/2 =
63.5\,\text{MPa}$, so the Tresca equivalent stress is $2\tau_{\max} =
127\,\text{MPa}$ and the Tresca safety factor is $250/127 = 1.97$.
Tresca is the more conservative of the two, as expected from its
inscribed hexagon. An engineer reporting a margin should state which
criterion was used; the same stress state is at safety factor $2.27$ by
von Mises and $1.97$ by Tresca. The computation matches
\repopath{code/von_mises.py}.

\subsection*{Worked example 4: stress concentration at a hole}

\textbf{Problem.} A wide steel plate carries a remote uniaxial tension
of $\sigma_\infty = 80\,\text{MPa}$ and contains a small circular hole.
What is the peak stress at the hole, and how far from the hole does the
disturbance decay to within five percent of the remote field?

\textbf{Solution.} The Kirsch solution gives a stress-concentration
factor of $K_t = 3$ at the sides of the hole (the points transverse to
the load, $\theta = 90^\circ$ in figure~\ref{fig:vol03ch09:kirsch-hole}),
so the peak hoop stress is
\[
\sigma_{\max} = K_t\,\sigma_\infty = 3\times 80 = 240\,\text{MPa},
\]
which exceeds the $250\,\text{MPa}$ yield by only a small margin; a
hole that triples the local stress turns a comfortably elastic member
into one on the edge of local yielding. The Kirsch field shows the
hoop stress along the transverse axis decaying as $\sigma_\theta/
\sigma_\infty = 1 + \tfrac{1}{2}(a/r)^2 + \tfrac{3}{2}(a/r)^4$ for
$r \ge a$. At $r = 3a$ this gives $1 + 0.056 + 0.0185 = 1.074$, still
seven percent high; at $r = 4a$ it is $1 + 0.031 + 0.0059 = 1.037$,
within four percent. The disturbance is local to within three or four
hole radii, the quantitative basis for the rule that a hole far from
an edge can be analysed as if the plate were infinite. A coarse FEM
mesh that places fewer than several elements within one radius of the
hole will miss this peak entirely.

\subsection*{Worked example 5: a shell-buckling margin}

\textbf{Problem.} A thin steel hemisphere of radius $R = 1\,\si{\meter}$
and wall thickness $h = 5\,\si{\milli\meter}$ ($E = 200\,\text{GPa}$,
$\nu = 0.3$) is subjected to external pressure. Find the linear-theory
critical pressure, apply a realistic imperfection knock-down, and state
the working buckling pressure.

\textbf{Solution.} The classical buckling pressure of a complete
spherical shell under external pressure is
\[
p_{\text{cr}} = \frac{2E h^2}{R^2\sqrt{3(1-\nu^2)}}
= \frac{2\times 200\times 10^9 \times (0.005)^2}
{1^2 \times \sqrt{3\times 0.91}}.
\]
The numerator is $2\times 200\times 10^9 \times 2.5\times 10^{-5} =
1.0\times 10^7$; the denominator is $\sqrt{2.73} = 1.652$, so
$p_{\text{cr}} = 6.05\times 10^6\,\text{Pa} = 6.05\,\text{MPa}$. This is
the perfect-shell value. Applying a knock-down factor of $0.3$, typical
for a sphere under external pressure (\repopath{data/shell_knockdown.csv}),
the working buckling pressure is $0.3\times 6.05 = 1.82\,\text{MPa}$, a
little under $18$ atmospheres. A vacuum failure ($\Delta p \le 0.1\,
\text{MPa}$) is nowhere near this margin, but a hemisphere that has to
resist several megapascals of external pressure would be designed
against the knocked-down value, never against the $6\,\text{MPa}$
linear figure. The factor-of-three penalty between linear theory and
working capacity is the price of imperfection sensitivity, and an
analyst who quotes the linear value as the capacity has overstated the
strength of the shell by a factor of three.

\subsection*{Worked example 6: a rectangular plate and the aspect-ratio effect}

\textbf{Problem.} Two simply supported steel plates carry the same
uniform pressure $q_0 = 4\,\text{kPa}$ and have the same thickness
$h = 8\,\si{\milli\meter}$ ($E = 200\,\text{GPa}$, $\nu = 0.3$). The
first is square, side $a = 1.2\,\si{\meter}$. The second is a long
rectangle, $a = 1.2\,\si{\meter}$ by $b = 2.4\,\si{\meter}$. Compare
the centre deflections and explain what the aspect ratio does to the
load path.

\textbf{Solution.} The flexural rigidity is common to both:
\[
D = \frac{E h^3}{12(1-\nu^2)} = \frac{200\times 10^9\times (0.008)^3}
{12\times 0.91} = 9.38\times 10^3\,\text{N}\cdot\text{m}.
\]
For a simply supported rectangular plate the centre deflection is
$w_{\max} = \alpha\, q_0 a^4/D$, where $a$ is the shorter side and the
coefficient $\alpha$ rises from $0.00406$ for a square toward the
one-way-strip limit $0.0130$ as the plate grows long. Tabulated values
give $\alpha = 0.00406$ at $b/a = 1$ and $\alpha = 0.01013$ at $b/a =
2$ (the collected coefficients are in
\repopath{data/plate_coefficients.csv}). With $a^4 = 1.2^4 = 2.074\,
\text{m}^4$ and $q_0 a^4/D = 4000\times 2.074/9380 = 0.884\,\text{m}$,
\[
w_{\text{square}} = 0.00406\times 0.884 = 3.59\times 10^{-3}\,\text{m}
\approx 3.6\,\si{\milli\meter},
\]
\[
w_{\text{long}} = 0.01013\times 0.884 = 8.96\times 10^{-3}\,\text{m}
\approx 9.0\,\si{\milli\meter}.
\]
The long plate deflects about $2.5$ times as much, although the short
span is identical. The square plate sends load to all four edges by
two-way bending, so each direction carries roughly half; the long
plate carries almost the whole load across the short span in one-way
bending, with the long edges contributing little once $b/a$ exceeds
about two. The lesson is the working rule that a rectangular plate of
aspect ratio above two behaves as a one-way strip spanning the short
direction, and is designed as such. The deflection-to-thickness ratios
are $0.45$ and $1.1$, so the long plate is at the edge of the
small-deflection regime and a more careful analysis would check whether
membrane stiffening has begun to help.

\subsection*{Worked example 7: a stiffened panel by smeared stiffness}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-stiffened-panel}

\textbf{Problem.} A steel deck panel, $2\,\si{\meter}\times 1\,
\si{\meter}$, is a $6\,\si{\milli\meter}$ skin reinforced by
longitudinal blade stiffeners $50\,\si{\milli\meter}$ deep and $6\,
\si{\milli\meter}$ thick at a pitch of $250\,\si{\milli\meter}$
(figure~\ref{fig:vol03ch09:stiffened-panel}). Estimate the effective
bending rigidity in the stiffener direction and compare it to the bare
skin, then state the separate check the smeared model cannot supply.

\textbf{Solution.} The bare-skin rigidity per unit width is $D_{\text{
skin}} = E h^3/[12(1-\nu^2)] = 200\times 10^9\times (0.006)^3/[12\times
0.91] = 3.96\times 10^3\,\text{N}\cdot\text{m}$. The stiffener adds
second moment of area along its length. Treat one stiffener plus its
tributary skin strip (width $b = 0.25\,\si{\meter}$) as a tee section.
The skin strip contributes area $A_s = 0.25\times 0.006 = 1.5\times
10^{-3}\,\text{m}^2$; the blade contributes $A_b = 0.05\times 0.006 =
3.0\times 10^{-4}\,\text{m}^2$. The neutral axis sits a distance
$\bar{y}$ below the skin midplane,
\[
\bar{y} = \frac{A_b\,(0.003 + 0.025)}{A_s + A_b}
= \frac{3.0\times 10^{-4}\times 0.028}{1.8\times 10^{-3}}
= 4.67\times 10^{-3}\,\text{m},
\]
about $4.7\,\si{\milli\meter}$ below the skin midplane. The section
second moment about that axis, summing the skin and blade own-inertias
and their parallel-axis terms, is approximately
\[
I = 1.5\times 10^{-3}\,(4.67\times 10^{-3})^2 + 3.0\times 10^{-4}\,
(0.0233)^2 + \tfrac{1}{12}(0.006)(0.05)^3 \approx 2.6\times 10^{-7}\,
\text{m}^4
\]
per stiffener bay. The smeared rigidity in the stiffener direction is
$D_x = E I/b = 200\times 10^9\times 2.6\times 10^{-7}/0.25 = 2.08\times
10^5\,\text{N}\cdot\text{m}$, about $53$ times the bare-skin rigidity.
The transverse rigidity $D_y$ stays near the bare-skin value, so the
panel is strongly orthotropic, $D_x \gg D_y$, and the global deflection
analysis uses the orthotropic plate equation $D_x w_{,xxxx} + 2H
w_{,xxyy} + D_y w_{,yyyy} = q$ rather than the isotropic biharmonic.

The check the smeared model cannot supply is the local one. Smearing
replaces the discrete skin-and-stiffener geometry with a continuum, so
it predicts the panel's global deflection and the stress carried into
the stiffeners, but it knows nothing about the skin bay between two
stiffeners. That bay, $250\,\si{\milli\meter}\times$ (its length),
$6\,\si{\milli\meter}$ thick, is a small plate in its own right and can
buckle locally under in-plane compression or deflect excessively under
pressure long before the global panel is in trouble. A complete design
runs the smeared model for the global response and a separate local
plate analysis for the worst skin bay. The smeared-versus-discrete
split is the plate-and-shell instance of the multiscale discipline:
one model for the structure, another for the detail.

\subsection*{Worked example 8: discontinuity stress at a shell junction}

\textbf{Problem.} A cylindrical shell, radius $R = 0.6\,\si{\meter}$,
wall $h = 12\,\si{\milli\meter}$, internal pressure $p = 3\,\text{MPa}$,
is closed by a hemispherical head of the same radius but wall $h/2 =
6\,\si{\milli\meter}$, the membrane-efficient thickness. Estimate the
radial growth mismatch at the junction and the axial length over which
the bending boundary layer decays.

\textbf{Solution.} Under pressure each shell grows radially. The
cylinder hoop strain is $\epsilon_\theta = (\sigma_\theta - \nu
\sigma_z)/E$ with $\sigma_\theta = pR/h$ and $\sigma_z = pR/2h$, so
\[
\delta_{\text{cyl}} = R\,\epsilon_\theta = \frac{R}{E}\,
\frac{pR}{h}\left(1 - \tfrac{\nu}{2}\right)
= \frac{0.6}{200\times 10^9}\,\frac{3\times 10^6\times 0.6}{0.012}\,
(1 - 0.15).
\]
The membrane hoop stress is $pR/h = 3\times 10^6\times 0.6/0.012 =
150\,\text{MPa}$, so $\delta_{\text{cyl}} = (0.6/200\times 10^9)\times
150\times 10^6\times 0.85 = 3.83\times 10^{-4}\,\text{m}$, about
$0.38\,\si{\milli\meter}$. The hemisphere carries equal biaxial stress
$pR/2h_{\text{sph}} = 3\times 10^6\times 0.6/(2\times 0.006) = 150\,
\text{MPa}$ as well (the half thickness exactly offsets the factor of
two), and its radial growth is $\delta_{\text{sph}} = (R/E)\,\sigma\,
(1-\nu) = (0.6/200\times 10^9)\times 150\times 10^6\times 0.7 = 3.15
\times 10^{-4}\,\text{m}$, about $0.32\,\si{\milli\meter}$.

The two free shells want to grow by different amounts, a mismatch of
about $0.07\,\si{\milli\meter}$, but the junction holds them together.
The shells satisfy continuity by developing shear and bending at the
edge, a self-equilibrating set of edge forces that produces the local
\engterm{discontinuity stress}. The bending decays over the
characteristic length of the shell boundary layer,
\[
\ell \approx \frac{\pi}{\beta}, \qquad
\beta = \left[\frac{3(1-\nu^2)}{R^2 h^2}\right]^{1/4},
\]
which for the cylinder gives $\beta = [3\times 0.91/(0.6^2\times
0.012^2)]^{1/4} = [2.73/(5.18\times 10^{-5})]^{1/4} = (5.27\times
10^4)^{1/4} = 15.1\,\text{m}^{-1}$, so the bending is confined to about
$1/\beta = 66\,\si{\milli\meter}$, of order $\sqrt{R h} = \sqrt{0.6
\times 0.012} = 0.085\,\si{\meter}$. The boundary-layer stress can add
a meaningful fraction of the membrane stress at the junction
(figure~\ref{fig:vol03ch09:discontinuity-stress}), which is why the
junction is detailed with a thickness taper or a knuckle radius and is
never sized on the membrane formula alone. The estimate sets the scale:
the local detail matters over a band a few times $\sqrt{R h}$ wide, and
nowhere else.

\subsection*{Worked example 9: a bolted flat circular cover}

\textbf{Problem.} A flat circular steel cover plate closes a
$0.5\,\si{\meter}$ inner-diameter access opening (radius $a = 0.25\,
\si{\meter}$) and is held by a ring of bolts on the rim. The chamber
behind it carries an internal gauge pressure $p = 1.2\,\text{MPa}$
($E = 200\,\text{GPa}$, $\nu = 0.3$, $\sigma_Y = 250\,\text{MPa}$).
Size the plate thickness against the bending stress, then size the bolt
ring against the separating force.

\textbf{Solution.} A flat cover carries pressure in bending, not in
membrane, which is why a flat closure is always heavier than a dished
head of the same opening. Treat the cover as a simply supported circular
plate (the bolt ring pins the rim but does not clamp it), for which the
maximum bending moment is at the centre,
\[
M_c = \frac{p a^2 (3+\nu)}{16}
= \frac{1.2\times 10^6 \times 0.25^2 \times 3.3}{16}
= 1.547\times 10^4\,\text{N}\cdot\text{m}/\text{m}.
\]
Setting the surface stress $\sigma = 6 M_c/h^2$ to a working allowable of
$\sigma_Y/1.67 = 150\,\text{MPa}$ and solving for the thickness gives
$h = \sqrt{6 M_c/\sigma_{\text{allow}}} = \sqrt{6\times 1.547\times
10^4/(150\times 10^6)} = 0.0249\,\si{\meter}$, so specify $h = 25\,
\si{\milli\meter}$. The realised centre stress at $25\,\si{\milli\meter}$
is $6\times 1.547\times 10^4/(0.025)^2 = 148.5\,\text{MPa}$, a yield
margin of $1.68$, and the centre deflection is $w_{\max} = p a^4(5+\nu)/
[64 D(1+\nu)] = 1.04\,\si{\milli\meter}$ on a flexural rigidity
$D = E h^3/[12(1-\nu^2)] = 2.86\times 10^5\,\text{N}\cdot\text{m}$, with
$w/h = 0.04$ well inside linear theory.

The bolts carry the total separating force, the pressure acting over the
full cover area,
\[
F = p\,\pi a^2 = 1.2\times 10^6 \times \pi \times 0.25^2 = 2.36\times
10^5\,\text{N} = 236\,\text{kN}.
\]
Sharing this over $N = 12$ bolts gives $F_b = 19.6\,\text{kN}$ per bolt.
An M12 grade-$8.8$ bolt has a tensile stress area of $84.3\,\si{\milli
\meter}^2$ and a proof strength near $580\,\text{MPa}$; the working
tensile stress is $19.6\times 10^3/(84.3\times 10^{-6}) = 233\,\text{
MPa}$, a margin of $2.5$ on proof, before any preload allowance. The
honest design preloads each bolt above the per-bolt service load so the
joint does not unload and leak under pressure, sizes the gasket and the
bolt circle so the cover seats uniformly, and checks the rim of the
plate for the local bending the bolt heads introduce. The flat cover is
simple to make and simple to seal, and it pays for that simplicity in
weight: a $2{:}1$ ellipsoidal head closing the same opening at the same
pressure would be a few millimetres thick, not twenty-five.

\subsection*{Worked example 10: the knuckle of a torispherical head}

\textbf{Problem.} A torispherical head closes a cylindrical vessel of
inner diameter $D_i = 1.2\,\si{\meter}$, wall $t = 10\,\si{\milli
\meter}$, under internal pressure $p = 1.6\,\text{MPa}$. The head has a
crown (spherical cap) radius $L = D_i = 1.2\,\si{\meter}$ and a knuckle
(toroidal transition) radius $r = 0.06 D_i = 72\,\si{\milli\meter}$.
Estimate the crown membrane stress and explain why the knuckle, not the
crown, governs the head design.

\textbf{Solution.} The crown is a spherical cap, so its membrane stress
is the sphere result with the crown radius in place of the vessel radius,
\[
\sigma_{\text{crown}} = \frac{p L}{2 t}
= \frac{1.6\times 10^6 \times 1.2}{2\times 0.010}
= 96\,\text{MPa},
\]
a benign tensile stress well under any steel allowable. The knuckle is
where the trouble sits. The toroidal knuckle turns the meridian from the
shallow crown back to the cylinder, and across that turn the
circumferential (hoop) membrane stress goes into \emph{compression},
because the surface is concave toward the pressure on the knuckle's inner
flank. A thin compressive hoop band in a shell is a buckling candidate,
and the membrane compression scales with the ratio of the crown radius
to the knuckle radius, here $L/2r = 1.2/0.144 = 8.3$, so the local stress
state in the knuckle is severe out of proportion to the gentle crown.
The knuckle also carries the bending boundary layer of the
crown-to-cylinder discontinuity, confined to a band of order $\sqrt{R t}
= \sqrt{0.6\times 0.010} = 77\,\si{\milli\meter}$, comparable to the
knuckle radius itself, so the membrane compression and the discontinuity
bending overlap in the same small region. Pressure-vessel codes size
torispherical heads against the knuckle, not the crown, and impose a
minimum knuckle radius (typically $6\,\%$ of the diameter, the value used
here) precisely to keep the compressive hoop and the local bending within
bounds. A head drawn with too tight a knuckle can buckle inward at the
knuckle under internal pressure, a failure that the crown membrane stress
of $96\,\text{MPa}$ gives no warning of whatever. The lesson carries
across the chapter: the membrane formula sizes the easy region and is
silent on the region that governs, which is always a discontinuity.

\subsection*{Worked example 11: thermal stress in a restrained plate}

\textbf{Problem.} A flat steel plate is part of a structure that holds
its edges fully against in-plane expansion. The plate is then heated
uniformly by $\Delta T = 80\,\si{\kelvin}$ above its installation temperature
($E = 200\,\text{GPa}$, $\nu = 0.3$, coefficient of thermal expansion
$\alpha = 12\times 10^{-6}\,\si{\per\kelvin}$, $\sigma_Y = 250\,\text{
MPa}$). Find the in-plane stress the restraint induces, and compare it to
the yield strength.

\textbf{Solution.} A free plate heated by $\Delta T$ would expand by the
thermal strain $\epsilon_{\text{th}} = \alpha\,\Delta T$ in every
in-plane direction, with no stress. Full in-plane restraint forces the
total strain to zero, so the mechanical strain must cancel the thermal
strain, $\epsilon_{\text{mech}} = -\alpha\,\Delta T$ in both directions.
For biaxial restraint the plane-stress Hooke law gives equal in-plane
stresses
\[
\sigma = -\frac{E\,\alpha\,\Delta T}{1-\nu}
= -\frac{200\times 10^9 \times 12\times 10^{-6}\times 80}{1-0.3}
= -2.74\times 10^8\,\text{Pa} = -274\,\text{MPa},
\]
a compressive stress that exceeds the $250\,\text{MPa}$ yield. A
restrained steel plate heated by only $80\,\si{\kelvin}$ yields in compression,
and on cooling it carries a residual tensile stress of similar
magnitude that can drive cracking or, in a thin plate, the
buckling-and-snap-through of thermal oil-canning. If the plate were
restrained in one direction only, the stress would be the smaller
uniaxial value $-E\,\alpha\,\Delta T = -192\,\text{MPa}$, still a large
fraction of yield. The factor $1/(1-\nu)$ between the uniaxial and
biaxial cases is the same Poisson amplification that makes plane-strain
constraint stiffen a material toward incompressibility, and it is easy to
forget when an analyst applies a one-dimensional intuition to a fully
restrained plate. The design responses are to provide expansion joints
or slip planes that relieve the restraint, to choose a low-expansion
material where restraint cannot be avoided, or to carry the thermal
stress explicitly as a load case in the yield and buckling checks. A
finite-element thermal-stress analysis that omits the restraint, or that
applies a free-expansion boundary where the real structure is held,
misses this entire stress field, which is one more entry in the
catalogue of boundary-condition errors of section 9.7.

\subsection*{Worked example 12: a clamped circular plate by superposition}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-clamped-superposition}

\textbf{Problem.} A circular steel plate of radius $a = 0.4\,\si{\meter}$
and thickness $h = 12\,\si{\milli\meter}$ ($E = 200\,\text{GPa}$,
$\nu = 0.3$) carries a uniform pressure $q_0 = 60\,\text{kPa}$. The rim
is clamped. Recover the clamped centre deflection by superposing the
simply supported plate under the same load with an edge moment, and show
that the answer agrees with the catalogue formula.

\textbf{Solution.} The clamped condition holds both the rim deflection
and the rim slope to zero. Build it from two solved axisymmetric cases.
The first is the simply supported plate under the same uniform load, for
which the rim deflection is zero but the rim rotates by
\[
\theta_{\text{ss}} = \frac{q_0 a^3}{8 D\,(1+\nu)},
\]
the standard simply supported edge-rotation result. The second is the
plate loaded by an edge moment $M_0$ alone, with no transverse load,
which produces a uniform curvature and a rim rotation
\[
\theta_{M} = \frac{M_0\, a}{D\,(1+\nu)}.
\]
The clamped condition is $\theta_{\text{ss}} + \theta_{M} = 0$, which
fixes the cancelling moment at $M_0 = -q_0 a^2/8$, the same edge moment
that the closed-form clamped solution carries at the rim
(figure~\ref{fig:vol03ch09:clamped-superposition}). The centre deflection
is the simply supported value reduced by the moment contribution. The
simply supported centre deflection is $w_{\text{ss}} = q_0 a^4 (5+\nu)/
[64 D (1+\nu)]$ and the uniform-moment plate lifts the centre by
$M_0 a^2/[2D(1+\nu)]$; combining gives the clamped centre deflection
$w_{\text{cl}} = q_0 a^4/(64 D)$, the catalogue result.

The numbers follow. The flexural rigidity is $D = E h^3/[12(1-\nu^2)] =
200\times 10^9 \times (0.012)^3/[12\times 0.91] = 3.16\times 10^4\,
\text{N}\cdot\text{m}$. The clamped centre deflection is
\[
w_{\text{cl}} = \frac{q_0 a^4}{64 D}
= \frac{60\times 10^3 \times (0.4)^4}{64\times 3.16\times 10^4}
= \frac{60\times 10^3 \times 0.0256}{2.02\times 10^6}
= 7.6\times 10^{-4}\,\text{m} = 0.76\,\si{\milli\meter}.
\]
The simply supported plate under the same load would deflect
$w_{\text{ss}} = (5+\nu)/[(1+\nu)] \times w_{\text{cl}} = (5.3/1.3)\times
0.76 = 3.1\,\si{\milli\meter}$, about four times as much, the
quantitative form of the rule that clamping a circular plate quarters
its centre deflection. The clamped plate pays for that with the rim
moment $M_0 = -q_0 a^2/8 = -60\times 10^3\times 0.16/8 = -1200\,\text{N}
\cdot\text{m}/\text{m}$, whose surface stress $6|M_0|/h^2 = 6\times 1200/
(0.012)^2 = 50\,\text{MPa}$ is the largest in the plate and sits at the
rim, not the centre. Superposition has turned two catalogue solutions and
one algebraic condition into the full clamped answer, and it gives every
edge restraint between simply supported and clamped by scaling the
cancelling moment between zero and $-q_0 a^2/8$.

\subsection*{Worked example 13: axisymmetric thermal stress in a disk}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-thermal-disk}

\textbf{Problem.} A solid steel disk of radius $a = 0.3\,\si{\meter}$ runs
with its centre hotter than its rim, the temperature falling parabolically
from a centre value $T_0 = 120\,\si{\kelvin}$ above the rim to zero at the
rim, $T(r) = T_0\,[1 - (r/a)^2]$ ($E = 200\,\text{GPa}$, $\alpha = 12\times
10^{-6}\,\si{\per\kelvin}$, $\sigma_Y = 250\,\text{MPa}$). Find the radial
and hoop stress distributions, locate the largest tensile stress, and
state which failure it drives.

\textbf{Solution.} A free disk with a non-uniform temperature carries no
net external load, but it cannot expand uniformly: the hot centre wants to
grow more than the cool rim allows, and the mismatch sets up a
self-equilibrating stress field. The axisymmetric thermoelastic solution
for a solid disk with a free rim gives the radial and hoop stresses in
terms of two temperature averages,
\[
\sigma_r = E\alpha\!\left[\frac{1}{a^2}\!\int_0^a T r\,dr
- \frac{1}{r^2}\!\int_0^r T r\,dr\right],
\quad
\sigma_\theta = E\alpha\!\left[\frac{1}{a^2}\!\int_0^a T r\,dr
+ \frac{1}{r^2}\!\int_0^r T r\,dr - T\right].
\]
For the parabolic profile the integrals are elementary. The disk-average
term is $(1/a^2)\int_0^a T r\,dr = T_0/4$, and the running average is
$(1/r^2)\int_0^r T r\,dr = T_0[\tfrac12 - \tfrac14 (r/a)^2]$. Substituting
collapses the result to the compact dimensionless form plotted in
figure~\ref{fig:vol03ch09:thermal-disk},
\[
\frac{\sigma_r}{E\alpha T_0} = \frac{1}{8}\!\left[\left(\tfrac{r}{a}
\right)^2 - 1\right],
\qquad
\frac{\sigma_\theta}{E\alpha T_0} = \frac{1}{8}\!\left[3\left(\tfrac{r}{a}
\right)^2 - 1\right].
\]
The radial stress is compressive everywhere inside and vanishes at the
free rim, as a traction-free edge demands. The hoop stress is compressive
near the hot centre, which is squeezed by the cooler ring around it, and
swings tensile beyond $r/a = 1/\sqrt 3$, reaching its largest value at the
rim,
\[
\sigma_{\theta}(a) = \tfrac{1}{4} E\alpha T_0
= \tfrac{1}{4}\times 200\times 10^9 \times 12\times 10^{-6}\times 120
= 7.2\times 10^7\,\text{Pa} = 72\,\text{MPa}.
\]
That $72\,\text{MPa}$ rim hoop tension is the crack-driving stress; the
disk cracks radially from the rim inward, not from the hot centre, which
is the observed failure of brake rotors and grinding wheels run too hot.
The margin against yield is $250/72 = 3.5$ for this temperature rise, but
the stress scales linearly with $T_0$, so a centre-to-rim difference of
about $430\,\si{\kelvin}$ would reach yield at the rim with no mechanical
load whatever. The lesson is the thermal-stress lesson of worked example
11 in axisymmetric form: a temperature field that is not free to expand
uniformly is a load, and the stress it produces is set by $E\alpha\Delta T$
and the geometry, not by any applied force.

\subsection*{Worked example 14: the fundamental frequency of a plate}

\textbf{Problem.} A simply supported square steel panel, side $a = 0.6\,
\si{\meter}$, thickness $h = 4\,\si{\milli\meter}$ ($E = 200\,\text{GPa}$,
$\nu = 0.3$, density $\rho = 7850\,\si{\kilogram\per\meter\cubed}$), is part
of a machine enclosure. Find its fundamental bending natural frequency and
say whether a $50\,\text{Hz}$ motor running nearby threatens resonance.

\textbf{Solution.} The free vibration of a Kirchhoff plate satisfies
$D\nabla^4 w + \rho h\, \ddot w = 0$, the plate bending equation with the
distributed load replaced by the inertia term $-\rho h\,\ddot w$. For a
simply supported rectangular plate the mode shapes are the same double
sines as the static Navier solution, $w_{mn} = \sin(m\pi x/a)\sin(n\pi y/b)$,
and substituting a mode $w_{mn}e^{i\omega t}$ turns the PDE into an
algebraic eigenvalue relation for the natural frequencies,
\[
\omega_{mn} = \pi^2\!\left[\left(\tfrac{m}{a}\right)^2
+ \left(\tfrac{n}{b}\right)^2\right]\sqrt{\frac{D}{\rho h}}.
\]
The fundamental is $m = n = 1$. The flexural rigidity is $D = E h^3/
[12(1-\nu^2)] = 200\times 10^9\times (0.004)^3/[12\times 0.91] = 1.17\times
10^3\,\text{N}\cdot\text{m}$, and the mass per area is $\rho h = 7850\times
0.004 = 31.4\,\si{\kilogram\per\meter\squared}$, so
\[
\sqrt{\frac{D}{\rho h}} = \sqrt{\frac{1.17\times 10^3}{31.4}}
= \sqrt{37.3} = 6.11\,\text{m}^2/\text{s}.
\]
With $a = b = 0.6\,\si{\meter}$, the bracket is $\pi^2[(1/0.6)^2 +
(1/0.6)^2] = \pi^2\times 2\times 2.78 = 54.8\,\text{m}^{-2}$, giving
$\omega_{11} = 54.8\times 6.11 = 335\,\text{rad/s}$ and
\[
f_{11} = \frac{\omega_{11}}{2\pi} = \frac{335}{6.28} = 53\,\text{Hz}.
\]
The panel's fundamental sits at $53\,\text{Hz}$, within a few percent of
the $50\,\text{Hz}$ excitation, so the motor does threaten resonance: the
panel would drum loudly and could fatigue at its welds. The cures all move
the frequency away from the excitation, and the frequency scales as
$h/a^2\sqrt{E/\rho}$, so the cheap move is a thicker panel or a stiffening
rib that shortens the effective span. Doubling the thickness to $8\,\si{
\milli\meter}$ doubles $f_{11}$ to about $106\,\text{Hz}$, clear of the
$50\,\text{Hz}$ source and its first harmonic. A static stress check would
have passed this panel without a hint of the resonance problem, because
resonance is a dynamic property the static analysis never computes.

\subsection*{Worked example 15: deflection of an orthotropic stiffened panel}

\textbf{Problem.} Take the stiffened deck panel of worked example 7,
$2\,\si{\meter}\times 1\,\si{\meter}$, with the smeared stiffener-direction
rigidity $D_x = 2.08\times 10^5\,\text{N}\cdot\text{m}$ already computed
there and a transverse rigidity near the bare-skin value $D_y = 3.96\times
10^3\,\text{N}\cdot\text{m}$. The panel is simply supported on all four
edges and carries a uniform pressure $q_0 = 20\,\text{kPa}$, with the
stiffeners running along the $2\,\si{\meter}$ direction. Estimate the
centre deflection from the orthotropic plate equation, and show what the
deflection would have been had the panel been treated as the bare
isotropic skin.

\textbf{Solution.} The orthotropic plate carries the load by the Huber
equation $D_x w_{,xxxx} + 2H w_{,xxyy} + D_y w_{,yyyy} = q$, with $H$ the
effective torsional rigidity. For a simply supported rectangular panel the
single-term Navier mode $w = w_0 \sin(\pi x/L_x)\sin(\pi y/L_y)$ satisfies
the equation term by term, and balancing it against a uniform load
(retaining the leading Fourier coefficient $16 q_0/\pi^2$) gives the centre
deflection
\[
w_0 = \frac{16 q_0/\pi^6}
{D_x/L_x^4 + 2H/(L_x^2 L_y^2) + D_y/L_y^4}.
\]
The stiffeners run along $x$, so the long $L_x = 2\,\si{\meter}$ pairs with
the large $D_x$ and the short $L_y = 1\,\si{\meter}$ pairs with the small
$D_y$. Evaluate the three denominator terms. The stiffener term is
$D_x/L_x^4 = 2.08\times 10^5/16 = 1.30\times 10^4$. The transverse term is
$D_y/L_y^4 = 3.96\times 10^3/1 = 3.96\times 10^3$. Taking the torsional
rigidity as the conservative geometric mean estimate $H \approx \sqrt{D_x
D_y} = \sqrt{2.08\times 10^5\times 3.96\times 10^3} = 2.87\times 10^4$, the
cross term is $2H/(L_x^2 L_y^2) = 2\times 2.87\times 10^4/(4\times 1) =
1.44\times 10^4$. The denominator sums to $1.30\times 10^4 + 1.44\times
10^4 + 3.96\times 10^3 = 3.14\times 10^4$. The numerator is $16\times 20
\times 10^3/\pi^6 = 3.20\times 10^5/961 = 333$. Hence
\[
w_0 = \frac{333}{3.14\times 10^4} = 1.06\times 10^{-2}\,\text{m}
\approx 11\,\si{\milli\meter}.
\]
Had the same panel been analysed as the bare $6\,\si{\milli\meter}$
isotropic skin (rigidity $D_{\text{skin}} = 3.96\times 10^3$ both ways),
the short span would govern and the simply supported deflection under
$20\,\text{kPa}$ would be of order $w \approx 0.00406\,q_0 L_y^4/D_{\text{
skin}} = 0.00406\times 20\times 10^3\times 1/3960 = 2.0\times 10^{-2}\,
\text{m} = 20\,\si{\milli\meter}$ if the bay spanned the full metre, and far
larger over the two-metre direction. The stiffeners roughly halve the
global deflection and reroute the load along their length, exactly the
orthotropy that the smeared rigidities encode. The result is a global
deflection only; the local skin-bay check of worked example 7 still has to
be run separately, because the smeared model knows nothing of the bay
between two stiffeners.

\subsection*{Worked example 16: a conical hopper holding liquid}

\textbf{Problem.} A steel conical hopper of half-angle $\alpha = 30^\circ$
hangs apex-down and holds a liquid of density $\rho = 1200\,\si{\kilogram
\per\meter\cubed}$ to a depth $H = 4\,\si{\meter}$ measured from the apex
along the axis. The wall thickness is $t = 6\,\si{\milli\meter}$ and the
material has $\sigma_Y = 250\,\text{MPa}$. Find the meridional and hoop
membrane stresses at the depth of greatest stress and locate that depth,
using the two membrane equations rather than a memorised formula.

\textbf{Solution.} The hopper is a shell of revolution, so the two
membrane equations of the shell section apply, but the load is now
hydrostatic and varies with depth rather than uniform. At an axial depth
$y$ below the apex the gauge pressure normal to the wall is the
hydrostatic head of the liquid above that level, $p(y) = \rho g (H - y)$
acting outward on the cone wall plus the head at the point itself; for a
hopper filled to $H$ the wall pressure at depth $y$ is $p(y) = \rho g
(H - y)$. The meridional resultant follows from axial balance on the cap
of liquid below the level $y$: the weight of liquid carried by the wall at
$y$ is the cone of liquid below, and dividing by the wall circumference
and the meridional direction cosine gives, for the conical geometry, a
meridional stress that rises from zero at the apex, peaks, and falls
toward the rim. Carrying the balance through for a cone, the meridional
stress is
\[
\sigma_\phi(y) = \frac{\rho g\, y\,(H - \tfrac{2}{3}y)}{2 t \cos\alpha} \cdot
\frac{\tan\alpha}{1},
\]
and the hoop stress, from the Laplace-Young equation with $r_1 \to \infty$
on the straight meridian and the hoop radius $r_2 = y\tan\alpha/\cos\alpha$,
is
\[
\sigma_\theta(y) = \frac{p(y)\, r_2}{t}
= \frac{\rho g (H - y)\, y \tan\alpha}{t \cos\alpha}.
\]
The hoop stress is the product of a head term $(H - y)$ that falls with
depth and a radius term $y$ that grows with depth, so it peaks where the
product is largest, at $y = H/2 = 2\,\si{\meter}$. There,
\[
\sigma_\theta = \frac{1200 \times 9.81 \times 2 \times 2 \times 0.577}
{0.006 \times 0.866}
= \frac{2.72\times 10^4}{5.20\times 10^{-3}}
= 5.2\times 10^{6}\,\text{Pa} \approx 5.2\,\text{MPa}.
\]
The hoop stress governs and peaks at mid-depth, not at the rim and not at
the apex, the result a constant-pressure intuition misses entirely. The
margin against yield is large here because the head is modest, but the
lesson is the location: a hydrostatically loaded cone is most heavily
stressed where the falling head and the growing radius cross, and a
designer who thickens the rim or the apex has reinforced the wrong band.
The same two equations that gave the uniform-pressure cone its linear
stress growth give the hydrostatic cone a peak in the middle, because the
load is no longer constant.

\subsection*{Worked example 17: axial buckling of a compressed cylinder}

\textbf{Problem.} A thin steel cylindrical shell, radius $R = 0.5\,\si{
\meter}$, wall $t = 3\,\si{\milli\meter}$ ($E = 200\,\text{GPa}$,
$\nu = 0.3$), carries an axial compressive load. Find the classical axial
buckling stress, apply the severe knock-down that axial compression
demands, and compare the working buckling stress to the material yield.

\textbf{Solution.} Axial compression is the most imperfection-sensitive of
the shell-buckling modes, the case figure~\ref{fig:vol03ch09:shell-buckling}
draws at the steepest knock-down. The classical critical axial stress of a
perfect cylinder is
\[
\sigma_{\text{cr}} = \frac{E\, t}{R\,\sqrt{3(1-\nu^2)}}
= \frac{200\times 10^9 \times 0.003}{0.5 \times \sqrt{2.73}}
= \frac{6.0\times 10^8}{0.5 \times 1.652}
= 7.26\times 10^{8}\,\text{Pa} = 726\,\text{MPa}.
\]
The perfect-shell value exceeds the $250\,\text{MPa}$ yield, which would
suggest the cylinder yields before it buckles. That conclusion is wrong,
and dangerously so, because the axial-compression knock-down is the most
brutal in the field. Test data put real cylinders at $15$ to $20\,\%$ of
the classical stress; taking the conventional $0.2$ factor (consistent
with \repopath{data/shell_knockdown.csv}), the working buckling stress is
\[
\sigma_{\text{work}} = 0.2 \times 726 = 145\,\text{MPa},
\]
now well below yield. The cylinder buckles before it yields, the reverse
of what the classical number implied. A designer who sized this shell on
the $726\,\text{MPa}$ classical stress, or who saw it exceed yield and
concluded buckling was not the governing mode, would overstate the
capacity by a factor approaching five. The discipline is the one the shell
section names: never the classical value, always the knocked-down value,
and for axial compression the knock-down is the largest of all.

\subsection*{Worked example 18: a foundation slab on elastic soil}

\textbf{Problem.} A square reinforced-concrete foundation slab, side
$L = 3\,\si{\meter}$, thickness $h = 0.30\,\si{\meter}$ ($E = 30\,\text{
GPa}$, $\nu = 0.2$), rests on soil with a modulus of subgrade reaction
$k = 20\,\si{\mega\pascal}/\si{\meter}$ (pressure per unit deflection). A
central column delivers a load $P = 600\,\text{kN}$. Estimate the
characteristic length over which the slab spreads the load into the soil,
and judge whether the slab acts as a rigid block or as a flexible plate.

\textbf{Solution.} A plate resting on a continuous elastic support is
governed by the plate-on-elastic-foundation equation, the
two-dimensional cousin of the beam on an elastic foundation of chapter 8,
\[
D\,\nabla^4 w + k\, w = q,
\]
where the soil reaction $k\,w$ replaces the free-edge condition with a
distributed restoring pressure. The competition between the plate's
flexural rigidity $D$ and the soil stiffness $k$ defines a
\engterm{characteristic length}
\[
\ell = \left(\frac{D}{k}\right)^{1/4},
\]
the radius over which a concentrated load is spread before the soil
reaction has absorbed it. The flexural rigidity is
\[
D = \frac{E h^3}{12(1-\nu^2)}
= \frac{30\times 10^9 \times (0.30)^3}{12\times (1-0.04)}
= \frac{30\times 10^9 \times 0.027}{11.52}
= 7.03\times 10^{7}\,\text{N}\cdot\text{m},
\]
so the characteristic length is
\[
\ell = \left(\frac{7.03\times 10^7}{20\times 10^6}\right)^{1/4}
= (3.52)^{1/4} = 1.37\,\si{\meter}.
\]
The slab half-width is $L/2 = 1.5\,\si{\meter}$, comparable to $\ell$,
which places this slab on the boundary between the two limiting regimes.
When $\ell$ is much larger than the slab, the slab is effectively rigid
and the soil pressure is nearly uniform; when $\ell$ is much smaller, the
load dies away within a patch and the slab edges carry almost nothing, the
flexible-plate limit. Here $\ell$ and the slab half-width are within a
factor of two, so the slab is neither rigid nor fully flexible: the
central pressure under the column is higher than the average bearing
pressure $P/L^2 = 600\times 10^3/9 = 67\,\text{kPa}$, and the edges carry
less. A rigid-block design that assumes uniform bearing under-reinforces
the slab centre. The honest analysis carries the plate-on-foundation
solution, and the characteristic length is the number that tells the
designer in advance which regime the slab is in before any detailed
calculation is run.

\subsection*{Worked example 19: a thick-walled cylinder by the Lam\'e solution}

\textbf{Problem.} A hydraulic cylinder of inner radius $a = 50\,\si{
\milli\meter}$ and outer radius $b = 80\,\si{\milli\meter}$ carries an
internal pressure $p = 60\,\text{MPa}$ with no external pressure
($\sigma_Y = 700\,\text{MPa}$). The wall is too thick for the membrane
formula. Find the hoop and radial stresses at the bore and at the outer
surface from the Lam\'e thick-wall solution, and compare the bore hoop
stress to the thin-wall estimate.

\textbf{Solution.} When the wall is not thin the stress varies through the
thickness and the membrane approximation $pR/t$ no longer holds. The
axisymmetric plane-strain elasticity problem (a long cylinder, away from
the ends) has the closed-form \engterm{Lam\'e solution}, in which the
radial and hoop stresses at radius $r$ are
\[
\sigma_r(r) = \frac{a^2 p}{b^2 - a^2}\left(1 - \frac{b^2}{r^2}\right),
\qquad
\sigma_\theta(r) = \frac{a^2 p}{b^2 - a^2}\left(1 + \frac{b^2}{r^2}\right).
\]
The constant prefactor is $a^2 p/(b^2 - a^2) = (0.050)^2 \times 60\times
10^6/[(0.080)^2 - (0.050)^2] = 2.5\times 10^{-3}\times 60\times 10^6/
(6.4 - 2.5)\times 10^{-3} = 1.5\times 10^5/3.9\times 10^{-3} = 3.85\times
10^7\,\text{Pa}$. At the bore $r = a$, $b^2/a^2 = (0.080/0.050)^2 =
2.56$, so
\[
\sigma_\theta(a) = 3.85\times 10^7 (1 + 2.56) = 1.37\times 10^8\,\text{Pa}
= 137\,\text{MPa}, \qquad \sigma_r(a) = -60\,\text{MPa},
\]
the radial stress at the bore being exactly the negative of the applied
pressure, as a loaded inner surface demands. At the outer surface
$r = b$, $b^2/r^2 = 1$, so the radial stress vanishes (a free outer
surface) and the hoop stress is $\sigma_\theta(b) = 2\times 3.85\times
10^7 = 77\,\text{MPa}$. The hoop stress is largest at the bore and falls
toward the outer wall, the opposite of the uniform stress a membrane model
would report. The thin-wall estimate uses the mean radius $R = 65\,\si{
\milli\meter}$ and thickness $t = 30\,\si{\milli\meter}$: $pR/t = 60\times
10^6\times 0.065/0.030 = 130\,\text{MPa}$, within five percent of the
Lam\'e bore value here, but the agreement degrades fast as the wall
thickens, and the thin-wall formula gives a single number where the real
stress varies by nearly a factor of two through the wall. The bore is the
critical location and the section to check against yield; the von Mises
stress there, with the axial stress for a closed cylinder added, is the
quantity the design factor is applied to.

\subsection*{Worked example 20: in-plane bending stress in a pipe elbow}

\textbf{Problem.} A long-radius pipe elbow of mean pipe radius
$R = 0.15\,\si{\meter}$, wall $t = 8\,\si{\milli\meter}$, and bend radius
$R_b = 0.45\,\si{\meter}$ carries internal pressure and an in-plane
bending moment. Estimate the flexibility factor and the stress-
intensification factor of the elbow relative to a straight pipe, and say
why an elbow is both more flexible and more highly stressed than the
straight pipe it joins.

\textbf{Solution.} A curved pipe under bending does not behave as a beam of
its nominal section, because the cross-section ovalises under the bending,
the same Brazier-type flattening the shell section names for a straight
tube. The ovalisation makes the elbow far more flexible than beam theory
predicts and concentrates the stress beyond the nominal bending stress.
The governing dimensionless group is the \engterm{pipe factor}
\[
\lambda = \frac{t\, R_b}{R^2}
= \frac{0.008 \times 0.45}{(0.15)^2}
= \frac{3.6\times 10^{-3}}{2.25\times 10^{-2}}
= 0.16.
\]
The von K\'arm\'an analysis of curved pipes gives a flexibility factor
$k \approx 1.65/\lambda$ for small $\lambda$, here
\[
k \approx \frac{1.65}{0.16} = 10.3,
\]
so the elbow is about ten times as flexible in bending as a straight pipe
of the same section, which is why piping flexibility analysis cannot treat
elbows as rigid corners: they act as built-in flexible joints that absorb
thermal expansion. The stress-intensification factor, the elbow peak
stress over the nominal bending stress, follows the companion relation
$i \approx 0.9/\lambda^{2/3}$, here
\[
i \approx \frac{0.9}{0.16^{2/3}} = \frac{0.9}{0.295} = 3.05,
\]
so the elbow carries roughly three times the nominal bending stress at its
most-stressed point. The two factors together state the elbow's character:
it relieves the line's thermal load by flexing (the flexibility factor),
and it pays for that flex with a local stress concentration (the
intensification factor). A piping designer who models elbows as rigid
corners both overstates the line's stiffness, predicting thermal stresses
that the real flexible elbows would relieve, and understates the local
elbow stress, missing the very point where the line is most likely to
crack. Both errors flow from ignoring the ovalisation that the pipe factor
$\lambda$ measures.

\subsection*{Worked example 21: in-plane buckling of a compressed plate}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-plate-buckling-coefficient}

\textbf{Problem.} A simply supported steel plate, thickness $h = 6\,\si{
\milli\meter}$, width $b = 0.5\,\si{\meter}$ across the load and length
$a = 2\,\si{\meter}$ along it ($E = 200\,\text{GPa}$, $\nu = 0.3$), carries
a uniform in-plane compression on its short edges. Find the critical
buckling stress, the compressive load it corresponds to, and compare that
load to the load that would yield the plate.

\textbf{Solution.} A plate compressed in its own plane buckles out of plane
at a stress well below the column-buckling stress of a strip of the same
slenderness, because the unloaded edges brace the plate and carry part of
the load around the buckle. The critical stress is
\[
\sigma_{\text{cr}} = k\,\frac{\pi^2 D}{b^2 h},
\qquad D = \frac{E h^3}{12(1-\nu^2)},
\]
with $k$ the buckling coefficient of
figure~\ref{fig:vol03ch09:plate-buckling-coefficient}. The plate is four
times as long as it is wide, so $a/b = 4$ sits on the flat part of the
envelope and $k = 4.0$. The flexural rigidity is $D = 200\times 10^9\times
(0.006)^3/[12\times 0.91] = 3.96\times 10^3\,\text{N}\cdot\text{m}$, so
\[
\sigma_{\text{cr}} = 4.0\times \frac{\pi^2 \times 3.96\times 10^3}
{(0.5)^2 \times 0.006}
= 4.0\times \frac{3.91\times 10^4}{1.5\times 10^{-3}}
= 1.04\times 10^{8}\,\text{Pa} = 104\,\text{MPa}.
\]
The plate buckles at $104\,\text{MPa}$, well below the $250\,\text{MPa}$
yield, so buckling governs and the plate never reaches its material
strength. The buckling load is $\sigma_{\text{cr}}\,b\,h = 104\times 10^6
\times 0.5\times 0.006 = 3.1\times 10^5\,\text{N} = 310\,\text{kN}$,
against a squash load $\sigma_Y b h = 750\,\text{kN}$ that the plate cannot
reach. The design lesson is the one the buckling-coefficient curve carries:
the critical stress scales as $(h/b)^2$, so the width across the load, not
the length, is the parameter to control, and a long compressed plate is
stiffened at a spacing near its width to force $k$ back up the steep part
of the curve rather than left to sit at the floor $k=4$.

\subsection*{Worked example 22: stress in a spinning disk}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-spinning-disk}

\textbf{Problem.} A solid steel disk of radius $a = 0.25\,\si{\meter}$ and
uniform thickness spins at $n = 6000$ revolutions per minute ($\rho =
7850\,\si{\kilogram\per\meter\cubed}$, $\nu = 0.3$, $\sigma_Y = 400\,
\text{MPa}$). Find the peak stress, locate it, and give the burst speed.

\textbf{Solution.} A spinning disk carries an inertial body force
$\rho\omega^2 r$ per unit volume, directed outward and growing with radius,
the centrifugal load. The plane-stress solution for a solid disk with a
free rim gives radial and hoop stresses that both peak at the centre, where
they are equal,
\[
\sigma_{\max} = \sigma_r(0) = \sigma_\theta(0)
= \frac{3+\nu}{8}\,\rho\,\omega^2 a^2
\qquad(\text{figure~\ref{fig:vol03ch09:spinning-disk}}).
\]
The angular speed is $\omega = 2\pi n/60 = 2\pi\times 6000/60 = 628\,
\text{rad/s}$, so $\omega^2 = 3.95\times 10^5\,\text{s}^{-2}$. With
$(3+\nu)/8 = 3.3/8 = 0.4125$,
\[
\sigma_{\max} = 0.4125\times 7850\times 3.95\times 10^5\times (0.25)^2
= 0.4125\times 7850\times 3.95\times 10^5\times 0.0625
= 8.0\times 10^{7}\,\text{Pa} = 80\,\text{MPa}.
\]
The peak is $80\,\text{MPa}$ at the axis, a margin of $5$ against the
$400\,\text{MPa}$ yield at this speed. The stress scales as $\omega^2$, so
the burst speed, where the centre stress reaches yield, is $\omega_{\text{
burst}} = \omega\sqrt{\sigma_Y/\sigma_{\max}} = 628\sqrt{400/80} = 628\times
2.24 = 1405\,\text{rad/s}$, about $13\,400$ revolutions per minute, a
factor of $\sqrt 5 \approx 2.2$ above the running speed. A central bore
removes the very material that carries the peak and nearly doubles the bore
hoop stress, which is why a bored flywheel or turbine disk bursts at a lower
speed than the solid disk this example sizes, and why rotating machinery is
spin-tested above its service speed but well below the calculated burst.

\subsection*{Worked example 23: curvature of a bimetallic strip}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-bimetallic-strip}

\textbf{Problem.} A bimetallic strip bonds a brass layer ($\alpha_1 =
19\times 10^{-6}\,\si{\per\kelvin}$) to an Invar layer ($\alpha_2 = 1.2
\times 10^{-6}\,\si{\per\kelvin}$), each of thickness $t = 0.3\,\si{\milli
\meter}$, total thickness $h = 0.6\,\si{\milli\meter}$, length $L = 40\,
\si{\milli\meter}$, both moduli near $E = 110\,\text{GPa}$. Find the tip
deflection of the strip, cantilevered at one end, for a temperature rise
$\Delta T = 50\,\si{\kelvin}$.

\textbf{Solution.} The two bonded layers want to grow by different amounts,
and because they cannot slide the strip curves to reconcile the mismatch,
the plate-and-shell instance of the thermal-load lesson of worked examples
11 and 13. For two layers of equal thickness and equal modulus the
curvature reduces to the Timoshenko bimetal result
\[
\kappa = \frac{3}{2}\,\frac{(\alpha_1 - \alpha_2)\,\Delta T}{h},
\]
independent of the modulus because the layers are matched. The expansion
mismatch is $\Delta\alpha = (19 - 1.2)\times 10^{-6} = 17.8\times 10^{-6}\,
\si{\per\kelvin}$, so
\[
\kappa = \frac{3}{2}\times \frac{17.8\times 10^{-6}\times 50}{6\times 10^{-4}}
= \frac{3}{2}\times \frac{8.9\times 10^{-4}}{6\times 10^{-4}}
= 2.23\,\text{m}^{-1}.
\]
The tip deflection of a cantilever bent to a uniform curvature is $\delta =
\kappa L^2/2 = 2.23\times (0.040)^2/2 = 1.78\times 10^{-3}\,\text{m} =
1.8\,\si{\milli\meter}$. A $40\,\si{\milli\meter}$ strip tip moves nearly
two millimetres for a $50\,\si{\kelvin}$ rise, a deflection large enough to
make or break a contact, which is the whole design of the thermostat and
the thermal cut-out. The deflection scales as $\Delta\alpha\,\Delta T\,
L^2/h$, so the sensitivity is maximised by pairing a high-expansion metal
with a near-zero-expansion one (brass and Invar here) and by making the
strip thin and long. The same bonded-mismatch mechanism, driven by cure
shrinkage rather than heat, warps an unbalanced composite laminate, and the
same formula sets the warp.

\subsection*{Worked example 24: the annular base plate of a column}

\textbf{Problem.} A steel column carries an axial load $P = 900\,\text{kN}$
onto a circular base plate of outer radius $b = 0.30\,\si{\meter}$ that
bears on a concrete foundation. The column is a tube of outer radius
$a = 0.12\,\si{\meter}$ that loads the plate around that circle. Estimate
the base-plate thickness so the bearing pressure stays within the concrete
allowable and the plate bending stress stays within the steel allowable
($\sigma_{\text{allow,steel}} = 165\,\text{MPa}$, concrete bearing allowable
$f_c = 14\,\text{MPa}$).

\textbf{Solution.} The base plate spreads a concentrated column load into a
distributed bearing pressure on the foundation, and it does so by bending,
so it is a circular plate loaded by a ring load at $r = a$ and supported by
the upward bearing pressure over its area. First the bearing check: the
mean pressure under the plate is
\[
q = \frac{P}{\pi b^2} = \frac{900\times 10^3}{\pi\times (0.30)^2}
= \frac{900\times 10^3}{0.283} = 3.18\times 10^{6}\,\text{Pa}
= 3.2\,\text{MPa},
\]
comfortably under the $14\,\text{MPa}$ concrete allowable, so the plan size
is adequate and bending sets the thickness. The plate cantilevers the
bearing pressure on the annulus outside the column from the loaded circle
$r = a$ out to the free rim $r = b$. Treating the overhang as a cantilever
strip of length $c = b - a = 0.18\,\si{\meter}$ carrying the uniform
bearing pressure $q$, the bending moment per unit width at the column face
is
\[
M = \frac{q\,c^2}{2} = \frac{3.18\times 10^6\times (0.18)^2}{2}
= 5.15\times 10^{4}\,\text{N}\cdot\text{m}/\text{m}.
\]
Setting the surface stress $6M/h^2$ to the steel allowable and solving,
\[
h = \sqrt{\frac{6M}{\sigma_{\text{allow}}}}
= \sqrt{\frac{6\times 5.15\times 10^4}{165\times 10^6}}
= \sqrt{1.87\times 10^{-3}} = 0.043\,\text{m},
\]
so specify a $45\,\si{\milli\meter}$ base plate. The cantilever-strip model
is the standard design idealisation for a column base plate, deliberately
conservative because it ignores the two-way plate action that would share
the load, and it is the method the steel-design codes encode. The thickness
is set by the overhang $c$ and the bearing pressure, not by the column load
directly, which is why a base plate only a little larger than the column is
thin while a widely oversized plate with a long overhang is thick.

\subsection*{Worked example 25: the pinched-cylinder benchmark}

\textbf{Problem.} A thin cylindrical shell is pinched by two equal and
opposite radial point loads $P$ at mid-length, a standard verification
benchmark for shell finite elements. State the reference radial deflection
under the load and explain what makes this problem a searching test of a
shell element, and how an analyst uses it before trusting a commercial code
on a real shell.

\textbf{Solution.} The pinched cylinder (radius $R$, length $L$, wall $t$,
rigid diaphragms at the ends, two opposed radial loads $P$ at the centre) is
one of the small set of shell benchmarks whose accepted answer is known to
high accuracy from series and fine-mesh solutions. The reference radial
deflection under each load, in the standard normalisation, is
\[
w_{\text{ref}} = 1.8248\times 10^{-5}\ \frac{P R}{E t^3}\times(\text{geometry
factors}),
\]
quoted for the canonical case $R/t = 100$, $L/R = 2$ as the fixed
dimensionless number the literature carries. The problem is searching for
three reasons that map onto the failure modes of section 9.7. The
deformation is bending-dominated with almost no membrane stretching under
the load, so an element that suffers shear or membrane locking reports a
deflection far too small, exactly the locking artifact of
figure~\ref{fig:vol03ch09:shear-locking}. The point load creates a sharp
local field that demands mesh refinement at the load, so the problem also
tests convergence at a concentration. And the doubly curved geometry with
rigid end diaphragms exercises the element's handling of both membrane and
bending states in one model. An analyst runs this benchmark, and its
companions (the Scordelis-Lo roof, the hemisphere with a point load), on
the exact element type and mesh density intended for the real work, and
accepts the code only when the benchmark converges to the reference number.
The benchmark is code verification in the sense of section 9.7: it confirms
the tool solves a known problem correctly before it is turned loose on an
unknown one, the shell-analysis counterpart of the patch test the
constant-strain triangle had to pass.

\subsection*{Worked example 26: a corrugated panel as an orthotropic plate}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-corrugated-panel}

\textbf{Problem.} A steel roofing sheet of base thickness $t = 0.7\,\si{
\milli\meter}$ is roll-formed into a trapezoidal corrugation of depth
$d = 35\,\si{\milli\meter}$ and pitch $p = 175\,\si{\milli\meter}$ ($E =
200\,\text{GPa}$). Estimate the effective bending rigidity along the
corrugations and compare it to the flat sheet, then state the direction in
which the panel is weak.

\textbf{Solution.} Folding a flat sheet into a corrugation moves the metal
away from the panel's neutral plane, so the second moment of area per unit
width in the corrugation direction rises steeply with the profile depth,
the same thickness-away-from-the-midsurface principle the sandwich panel
uses. For a trapezoidal profile the second moment of area per unit width is
of order
\[
I_x \approx \frac{t\,d^2}{2}
= \frac{0.7\times 10^{-3}\times (0.035)^2}{2}
= 4.3\times 10^{-7}\,\text{m}^3/\text{m},
\]
against the flat-sheet value $I_{\text{flat}} = t^3/12 = (0.7\times
10^{-3})^3/12 = 2.9\times 10^{-11}\,\text{m}^3/\text{m}$. The corrugation
is stiffer in bending along the folds by the ratio $I_x/I_{\text{flat}}
\approx 1.5\times 10^4$, four orders of magnitude, so the effective rigidity
$D_x = E I_x = 200\times 10^9\times 4.3\times 10^{-7} = 8.6\times 10^{4}\,
\text{N}\cdot\text{m}$ against the flat-sheet $D_{\text{flat}} = 5.7\,
\text{N}\cdot\text{m}$. That is why a thin roofing sheet spans metres
between purlins along its corrugations without sagging. The panel is weak
across the corrugations: bending the sheet transverse to the folds only
flexes the thin flats and unfolds the profile, so the transverse rigidity
$D_y$ is close to the flat-sheet value and the panel is strongly
orthotropic, $D_x \gg D_y$. The corrugations are laid running along the
span the panel must carry, and a load applied across the folds finds almost
none of the stiffness the along-fold direction advertises. The corrugated
sheet is the cheapest orthotropic plate in the built environment, and it
carries the same local caution as the stiffened and sandwich panels: the
thin walls of an individual fold can buckle before the panel as a whole is
in trouble.

\subsection*{Worked example 27: the shear-correction of a thick plate}

\textbf{Problem.} A steel plate strip of span $L = 0.4\,\si{\meter}$ and
thickness $h = 40\,\si{\milli\meter}$ (a span-to-thickness ratio of only
$10$) is simply supported and carries a central line load. Estimate how
much the transverse shear adds to the deflection that Kirchhoff theory
alone predicts, and say whether the thin-plate theory is adequate.

\textbf{Solution.} Kirchhoff plate theory neglects transverse shear, which
is accurate only while the plate is thin ($h \le L/20$). This plate is at
$h = L/10$, twice as thick as the working limit, so the shear contribution
has to be checked. The Mindlin-Reissner theory carries the load by bending
and transverse shear in series, so the total deflection is the sum of a
bending part and a shear part,
\[
w = w_{\text{bend}} + w_{\text{shear}},
\qquad
\frac{w_{\text{shear}}}{w_{\text{bend}}} \sim
c\left(\frac{h}{L}\right)^2\frac{E}{\kappa G},
\]
with $\kappa = 5/6$ the shear-correction factor and $c$ an order-one
constant that depends on the support and load. For steel $E/G = 2(1+\nu) =
2.6$, so $E/(\kappa G) = 2.6/0.833 = 3.1$, and with $(h/L)^2 = 0.01$ and a
representative $c \approx 2.4$ for a simply supported strip under a central
load,
\[
\frac{w_{\text{shear}}}{w_{\text{bend}}} \approx 2.4\times 0.01\times 3.1
= 0.075.
\]
The shear adds about eight percent to the deflection at this span-to-depth
ratio, small but no longer negligible, and it grows as $(h/L)^2$, so a plate
at $h = L/5$ would carry a shear contribution near a third. The Kirchhoff
estimate is a mild underestimate here and would be a serious one for a
thicker plate or a sandwich panel with a shear-soft core, where the shear
term can exceed the bending term. The working rule follows from the
$(h/L)^2$ scaling: below $h = L/20$ the shear correction is under two
percent and Kirchhoff theory suffices; above it, the plate is analysed with
a shear-deformable theory or with shell elements that carry the correction
internally, the finite-element move the shear-deformable mastery box below
describes.

\subsection*{Worked example 28: a nozzle opening and area replacement}

\textbf{Problem.} A cylindrical vessel wall of required thickness $t_r =
9\,\si{\milli\meter}$ (specified $12\,\si{\milli\meter}$) is pierced by a
radial nozzle of inner radius $r_n = 60\,\si{\milli\meter}$ whose neck wall
is $8\,\si{\milli\meter}$ thick. Work the area-replacement check that a
pressure-vessel code requires, and say whether a reinforcing pad is needed.

\textbf{Solution.} Cutting a hole in a pressurised shell removes
load-carrying metal and concentrates the stress at the opening, the
stress-concentration mechanism of worked example 4 applied to a real
nozzle. The code rule is area replacement: the metal removed by the
opening, within a defined reinforcement zone around it, must be made up by
excess metal in the shell, the nozzle neck, and any added pad. The area
removed is the required shell thickness times the hole diameter,
\[
A_{\text{rem}} = 2 r_n\,t_r = 0.120\times 9\times 10^{-3}
= 1.08\times 10^{-3}\,\text{m}^2.
\]
The reinforcement available comes first from the excess shell wall over the
reinforcement zone, a width conventionally taken as the hole diameter
$2 r_n$ each side of the opening. The shell carries $12 - 9 = 3\,\si{
\milli\meter}$ of excess over the required, across a zone width of about
$2\times 2 r_n = 0.24\,\si{\meter}$, giving
\[
A_{\text{shell}} = (t - t_r)\times 2\times 2 r_n
= 3\times 10^{-3}\times 0.24 = 7.2\times 10^{-4}\,\text{m}^2.
\]
The nozzle neck contributes its own excess wall over the height it projects,
roughly $A_{\text{neck}} \approx 2\times (t_n - t_{nr})\,\times 2.5\sqrt{r_n
t_n}$; with a neck required thickness near $2\,\si{\milli\meter}$ and the
$8\,\si{\milli\meter}$ neck, the excess is $6\,\si{\milli\meter}$ over a
projection of about $2.5\sqrt{0.06\times 0.008} = 5.5\,\si{\centi\meter}$
each side, giving roughly $A_{\text{neck}} \approx 6\times 10^{-3}\times
0.11 = 6.6\times 10^{-4}\,\text{m}^2$. The available reinforcement
$A_{\text{shell}} + A_{\text{neck}} = 7.2\times 10^{-4} + 6.6\times 10^{-4}
= 1.38\times 10^{-3}\,\text{m}^2$ exceeds the required $1.08\times 10^{-3}\,
\text{m}^2$, so the opening is self-reinforced by the excess wall and neck
and no pad is needed. Had the sum fallen short, a reinforcing pad welded
around the nozzle would supply the deficit. The check is the working form of
the stress-concentration discipline: the code does not compute the peak
stress at the hole, it restores the removed load path so the opening does
not lower the vessel's pressure capacity.

\subsection*{Worked example 29: a shrink-fit interference by the Lam\'e solution}

\textbf{Problem.} A steel hub of outer radius $c = 90\,\si{\milli\meter}$ is
shrink-fitted onto a solid steel shaft of nominal radius $b = 50\,\si{
\milli\meter}$ with a radial interference $\delta = 40\,\si{\micro\meter}$
($E = 200\,\text{GPa}$, $\nu = 0.3$). Find the interface contact pressure
and the peak hoop stress in the hub, using the thick-wall Lam\'e solution.

\textbf{Solution.} A shrink fit is a controlled interference: the hub bore
is machined slightly smaller than the shaft, and forcing them together sets
up a contact pressure that grips the shaft, the same thick-wall elasticity
as the pressurised cylinder of worked example 19 with the pressure now
generated by the geometric misfit rather than a fluid. The radial
interference is shared between the hub, which is pushed out, and the shaft,
which is squeezed in, and the interface pressure $p_c$ follows from matching
the total radial closure to $\delta$. For a solid shaft inside a hub of
outer radius $c$, both of the same material, the relation is
\[
\delta = \frac{p_c\, b}{E}\left[\frac{c^2 + b^2}{c^2 - b^2} + \nu\right]
+ \frac{p_c\, b}{E}\left[1 - \nu\right]
= \frac{p_c\, b}{E}\left[\frac{c^2 + b^2}{c^2 - b^2} + 1\right],
\]
the first bracket the hub's outward growth at its bore, the second the
shaft's inward squeeze. With $c^2 = 8.1\times 10^{-3}$, $b^2 = 2.5\times
10^{-3}\,\text{m}^2$, the ratio $(c^2+b^2)/(c^2-b^2) = 10.6\times 10^{-3}/
5.6\times 10^{-3} = 1.89$, so the bracket is $1.89 + 1 = 2.89$. Solving for
the contact pressure,
\[
p_c = \frac{\delta\,E}{b\,(2.89)}
= \frac{40\times 10^{-6}\times 200\times 10^9}{0.050\times 2.89}
= \frac{8.0\times 10^{6}}{0.1445}
= 5.5\times 10^{7}\,\text{Pa} = 55\,\text{MPa}.
\]
The interface carries a $55\,\text{MPa}$ contact pressure, and that pressure
loads the hub as an internal pressure on a thick ring, so the hub's peak
hoop stress at its bore is the Lam\'e result $\sigma_\theta(b) = p_c\,
(c^2+b^2)/(c^2-b^2) = 55\times 1.89 = 104\,\text{MPa}$, tensile at the bore
and the location to check against yield. The grip torque the fit can
transmit is the contact pressure times the interface area times the
friction coefficient times the shaft radius, so the same $55\,\text{MPa}$
sizes the joint's torque capacity. The design walks a line: too little
interference and the fit slips, too much and the hub bore yields, and the
Lam\'e solution is what places the interference between those bounds.

\subsection*{Worked example 30: subsurface shear in a Hertzian contact}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-hertz-contact}

\textbf{Problem.} A steel railway wheel presses on a steel rail through a
contact patch that carries a normal load $P = 100\,\text{kN}$ over a line
contact of length $\ell = 12\,\si{\milli\meter}$, with an effective
contact half-width $b = 6\,\si{\milli\meter}$ ($E = 200\,\text{GPa}$,
$\nu = 0.3$, $\sigma_Y = 700\,\text{MPa}$ for the hardened rail steel).
Estimate the peak contact pressure, locate the point of maximum shear
stress, and say why rolling-contact fatigue spalls the rail from below its
surface rather than from the surface itself.

\textbf{Solution.} A localised contact does not spread its load through a
membrane or a plate; it drives a three-dimensional elastic stress field
into the half-space beneath the patch, and the governing analysis is the
Hertz contact theory, a specialisation of the three-dimensional elasticity
of the chapter's parent section to two bodies pressed together. For a line
contact the pressure distribution across the patch is elliptical, peaking at
the centre at
\[
p_{\max} = \frac{2 P}{\pi\,\ell\,b}
= \frac{2\times 100\times 10^3}{\pi\times 0.012\times 0.006}
= \frac{2.0\times 10^5}{2.26\times 10^{-4}}
= 8.8\times 10^{8}\,\text{Pa} = 880\,\text{MPa},
\]
larger than the yield strength read as a uniaxial number, which would wrongly
suggest instant failure. The contact is not uniaxial. Directly under the
patch the material is in triaxial compression, and a triaxial pressure does
not yield a metal, because yield is set by shear, not by hydrostatic
pressure. The quantity that matters is the maximum shear stress, and the
Hertz field puts its largest value not at the surface but buried below it, at
a depth of about $0.78\,b$ for a line contact, where it reaches
$\tau_{\max}\approx 0.30\,p_{\max}$
(figure~\ref{fig:vol03ch09:hertz-contact}). Here
\[
\tau_{\max} \approx 0.30\times 880 = 264\,\text{MPa},
\]
at a depth of about $0.78\times 6 = 4.7\,\si{\milli\meter}$. Comparing this
to the shear yield $\tau_Y = \sigma_Y/2 = 350\,\text{MPa}$ gives a margin of
$350/264 = 1.3$ against first yield, and that yield, when it comes, begins at
the buried point. Under the millions of wheel passes a rail sees, the
repeated buried shear reversal initiates a fatigue crack below the running
surface, which grows parallel to the surface and eventually lifts a flake of
metal off: the shelling and squat defects that rail inspection hunts for. The
lesson is the one the von Mises section makes in the contact setting: the
surface pressure is the wrong number to compare against yield, the buried
shear is the right one, and a rolling contact fails from the inside out.

\subsection*{Worked example 31: facesheet wrinkling of a sandwich panel}

\textbf{Problem.} A sandwich panel has aluminium facesheets of thickness
$t_f = 0.5\,\si{\milli\meter}$ ($E_f = 70\,\text{GPa}$) bonded to a foam core
of Young's modulus $E_c = 60\,\text{MPa}$ and shear modulus $G_c = 22\,
\text{MPa}$. The panel is loaded in edgewise compression. Estimate the
facesheet wrinkling stress and compare it to the facesheet material yield of
$250\,\text{MPa}$, then state why the global bending check of worked example
7 would never flag this mode.

\textbf{Solution.} A sandwich facesheet in compression can buckle into short
wavelengths on its own, riding on the elastic core beneath it, at a stress
far below the material yield and far below the panel's global buckling load.
The facesheet acts as a plate on an elastic foundation, the core supplying
the foundation stiffness, and the short-wavelength instability is
\engterm{wrinkling}. The classical wrinkling stress for a facesheet on a
thick core is
\[
\sigma_{\text{wr}} \approx 0.5\,\bigl(E_f\,E_c\,G_c\bigr)^{1/3},
\]
a cube-root blend of the three stiffnesses that reflects the facesheet
bending, the core through-thickness stretch, and the core shear all resisting
the wrinkle together. With the numbers,
\[
\sigma_{\text{wr}} \approx 0.5\,\bigl(70\times 10^9\times 60\times 10^6
\times 22\times 10^6\bigr)^{1/3}.
\]
The product inside is $70\times 60\times 22\times 10^{9+6+6} = 9.24\times
10^4\times 10^{21} = 9.24\times 10^{25}$, whose cube root is
$(9.24\times 10^{25})^{1/3} = (92.4)^{1/3}\times 10^{8} = 4.52\times 10^{8}$,
so
\[
\sigma_{\text{wr}} \approx 0.5\times 4.52\times 10^{8} = 2.3\times 10^{8}\,
\text{Pa} = 226\,\text{MPa}.
\]
The wrinkling stress of $226\,\text{MPa}$ sits just below the $250\,\text{MPa}$
facesheet yield, so the facesheet wrinkles fractionally before it yields, and
a panel designed on the yield stress alone would be unconservative by that
margin. The global bending check of worked example 7 treats the panel as a
smeared plate with an effective rigidity and reports the deflection and the
mean facesheet stress; it carries no information about the wavelength-scale
instability of the compression face, because smearing has erased the very
length scale on which the wrinkle lives. A sandwich design therefore runs the
global bending or buckling check and, separately, the facesheet wrinkling
check, the core shear-and-crush check, and the bond check, each on the local
core and adhesive properties. The panel that passes on stiffness fails on the
mode the stiffness check never examined, the same multiscale caution the
stiffened panel taught in discrete form and the sandwich section stated in
prose.

\subsection*{Worked example 32: effective width of a post-buckled plate}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-effective-width}

\textbf{Problem.} A simply supported steel plate of width $b = 300\,\si{
\milli\meter}$ and thickness $t = 3\,\si{\milli\meter}$ ($E = 200\,\text{GPa}$,
$\nu = 0.3$, $\sigma_Y = 250\,\text{MPa}$) is a compression flange in a
box-girder skin. It buckles well before it yields. Find the buckling stress,
then use the von K\'arm\'an effective-width rule to estimate the load the
plate still carries when its braced edges reach yield, and compare that to
the load at first buckling.

\textbf{Solution.} A plate does not collapse when it buckles. Unlike a
column, whose buckling is the end of its useful load, a plate braced along
its unloaded edges sheds load from its buckled middle to its still-straight
edges and keeps carrying more total load as the edge stress climbs
(figure~\ref{fig:vol03ch09:effective-width}). The buckling stress is the
plate-buckling result of worked example 21 with the simply supported
coefficient $k = 4.0$,
\[
\sigma_{\text{cr}} = k\,\frac{\pi^2 D}{b^2 t},
\qquad D = \frac{E t^3}{12(1-\nu^2)} = \frac{200\times 10^9\times (0.003)^3}
{12\times 0.91} = 4.95\times 10^2\,\text{N}\cdot\text{m},
\]
so
\[
\sigma_{\text{cr}} = 4.0\times \frac{\pi^2\times 4.95\times 10^2}
{(0.3)^2\times 0.003}
= 4.0\times \frac{4.89\times 10^3}{2.7\times 10^{-4}}
= 7.2\times 10^{7}\,\text{Pa} = 72\,\text{MPa}.
\]
The plate buckles at $72\,\text{MPa}$, well below yield. The load at first
buckling, taken over the full width, is $P_{\text{cr}} = \sigma_{\text{cr}}\,
b\,t = 72\times 10^6\times 0.3\times 0.003 = 6.5\times 10^4\,\text{N} =
65\,\text{kN}$. Past buckling the plate carries load only over an effective
width $b_e$ concentrated at the two edges, and the von K\'arm\'an rule sets
that width from the ratio of the buckling stress to the current edge stress,
\[
\frac{b_e}{b} = \sqrt{\frac{\sigma_{\text{cr}}}{\sigma_{\text{edge}}}}.
\]
When the braced edges reach yield, $\sigma_{\text{edge}} = \sigma_Y =
250\,\text{MPa}$, so
\[
\frac{b_e}{b} = \sqrt{\frac{72}{250}} = \sqrt{0.288} = 0.537,
\qquad b_e = 0.537\times 300 = 161\,\si{\milli\meter}.
\]
The ultimate load the plate carries is the yield stress over the effective
width,
\[
P_{\text{ult}} = \sigma_Y\,b_e\,t
= 250\times 10^6\times 0.161\times 0.003 = 1.2\times 10^{5}\,\text{N}
= 121\,\text{kN},
\]
almost double the first-buckling load of $65\,\text{kN}$. The plate carries
$86\,\%$ more load after it has buckled than at the instant it buckled, which
is why thin-walled steel and aluminium structures are routinely designed to
buckle in service and to draw on this post-buckling reserve, the effective
width shrinking as the edge stress rises. A column has no such reserve and a
shell has a negative one (it sheds load catastrophically); the plate is the
structural element whose post-buckling path is stable enough to bank on, and
the effective-width rule is how a designer banks it.

\subsection*{Worked example 33: a barrel-vault shell roof by membrane theory}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-barrel-vault}

\textbf{Problem.} A reinforced-concrete cylindrical (barrel) shell roof of
radius $R = 8\,\si{\meter}$, thickness $t = 80\,\si{\milli\meter}$, spans a
half-angle $\phi_0 = 40^\circ$ each side of the crown and a longitudinal
distance $L = 24\,\si{\meter}$ between end diaphragms, carrying its own dead
weight $g = 3.0\,\text{kPa}$ of surface area (concrete density folded in).
Estimate the transverse (arch-action) membrane force at the crown and the
longitudinal (beam-action) stress at mid-span, and say why the free edges
need edge beams.

\textbf{Solution.} A barrel-vault roof works two ways at once, and the
membrane theory of shells of the chapter's shell section handles both
together. Across its curvature the shell behaves as an arch, carrying the
weight by a transverse membrane force; along its length it behaves as a deep
beam spanning between the end diaphragms, carrying the weight by a
longitudinal membrane force that varies over the section like a beam bending
stress. Take the arch action first. A strip of the shell one unit long carries
its own weight as an arch of radius $R$, and near the crown the transverse
membrane compression per unit length is of order
\[
N_\phi \approx g\,R = 3.0\times 10^3\times 8 = 2.4\times 10^{4}\,\text{N}/
\text{m},
\]
a compression of $24\,\text{kN}$ per metre, so the transverse membrane stress
is $N_\phi/t = 2.4\times 10^4/0.08 = 3.0\times 10^{5}\,\text{Pa} = 0.30\,
\text{MPa}$, a benign compression for concrete. The beam action is the larger
effect and the one that governs. The shell spans $L = 24\,\si{\meter}$ as a
curved beam whose depth is the rise of the vault, $d = R(1-\cos\phi_0) =
8\times(1-\cos 40^\circ) = 8\times 0.234 = 1.87\,\si{\meter}$. The dead load
per unit length of span is the weight of one full arch, roughly $w = g\times
(2 R \phi_0) = 3.0\times 10^3\times (2\times 8\times 0.698) = 3.35\times 10^4\,
\text{N}/\text{m}$. The mid-span bending moment of a simply supported span is
$M = w L^2/8 = 3.35\times 10^4\times 24^2/8 = 2.41\times 10^{6}\,\text{N}\cdot
\text{m}$, and the longitudinal membrane stress at the crown, the compression
edge of the beam, is of order $M/(A_{\text{eff}}\,d)$; with an effective flange
area at the crown of about $A_{\text{eff}} \approx t\times(2 R \phi_0) =
0.08\times 11.2 = 0.90\,\text{m}^2$, the crown compressive stress is
\[
\sigma_{\text{long}} \approx \frac{M}{A_{\text{eff}}\,d}
= \frac{2.41\times 10^6}{0.90\times 1.87}
= 1.43\times 10^{6}\,\text{Pa} = 1.4\,\text{MPa},
\]
a modest compression at the crown, the top of the beam. The trouble is at the
free straight edges, the bottom of the beam, where the same beam action puts
the longitudinal membrane force into tension and the arch action, having no
support beyond the edge, pulls the shell inward. Pure membrane theory predicts
a tension the thin shell cannot carry without cracking, which is why every
barrel-vault roof carries an edge beam along its free edges
(figure~\ref{fig:vol03ch09:barrel-vault}), a stiff longitudinal member that
absorbs the edge tension and anchors the shell, the same role the wind girder
plays for the storage tank. The membrane analysis sizes the shell interior in
two lines of arch-and-beam reasoning; the edge, once again, is the
discontinuity the membrane formula cannot close, and the edge beam is the fix.

\subsection*{Worked example 34: fatigue life of a plate with a hole}

\textbf{Problem.} The plate-with-a-hole of worked example 4 (remote stress
$\sigma_\infty = 80\,\text{MPa}$, stress-concentration factor $K_t = 3$,
steel with an endurance limit $\sigma_e = 200\,\text{MPa}$ at $10^6$ cycles
and an ultimate strength $\sigma_u = 500\,\text{MPa}$) now carries a
fully-reversed cyclic remote stress of amplitude $\sigma_a = 80\,\text{MPa}$
rather than a static load. Estimate the fatigue life at the hole, and show how
much the fatigue notch sensitivity softens the static concentration factor.

\textbf{Solution.} A static stress concentration and a fatigue stress
concentration are not the same number. The static peak at the hole is
$K_t\sigma_\infty = 3\times 80 = 240\,\text{MPa}$, the calculation of worked
example 4. Under cyclic loading, though, the material feels a slightly smaller
effective concentration, because the sharp stress gradient at the hole means
only a small volume of material sees the peak, and fatigue crack initiation
responds to the stress averaged over a small process zone rather than to the
mathematical point value. The \engterm{fatigue notch factor} $K_f$ is related
to the static $K_t$ through a notch-sensitivity factor $q$,
\[
K_f = 1 + q\,(K_t - 1),
\]
with $q$ between zero (fully insensitive, $K_f = 1$) and one (fully sensitive,
$K_f = K_t$). For a steel of this strength with a hole radius of a few
millimetres, $q \approx 0.85$ is representative, so
\[
K_f = 1 + 0.85\times(3 - 1) = 2.7,
\]
a little below the static $K_t = 3$. The effective cyclic stress amplitude at
the hole is
\[
\sigma_{a,\text{local}} = K_f\,\sigma_a = 2.7\times 80 = 216\,\text{MPa}.
\]
This local amplitude exceeds the endurance limit of $200\,\text{MPa}$, so the
detail does not have infinite life; it sits in the finite-life regime and will
crack after a countable number of cycles. Estimating the life from the
high-cycle (Basquin) line between the ultimate strength at one cycle and the
endurance limit at $10^6$ cycles, on a log-log $S$-$N$ fit the life at a local
amplitude of $216\,\text{MPa}$ is a little short of $10^6$ cycles, of order a
few hundred thousand cycles. The design responses are the ones the
stress-concentration discipline names: enlarge the hole radius or add a
generous fillet to lower $K_t$, cold-work the hole to leave a compressive
residual stress that subtracts from the cyclic tension, or lower the remote
stress by thickening the plate. The example carries worked example 4 one step
further, from the static peak that a monotonic load cares about to the fatigue
peak that a cyclic load cares about, and it is the fatigue peak that governs
the Comet cabin cutout of the failure section: a static check on the cutout
passed, and the aircraft still cracked, because the load was cyclic and the
governing number was $K_f\sigma_a$, not the yield margin.

\subsection*{Worked example 35: thermal-shock fracture of a brittle plate}

\textbf{Problem.} A soda-lime glass plate ($E = 70\,\text{GPa}$, $\nu = 0.22$,
coefficient of thermal expansion $\alpha = 9\times 10^{-6}\,\si{\per\kelvin}$,
tensile fracture strength $\sigma_f = 50\,\text{MPa}$, thermal conductivity
$\kappa = 1.0\,\si{\watt\per\meter\per\kelvin}$) has one face suddenly cooled
while the other stays hot, as when a hot dish meets cold water. Estimate the
surface temperature difference the plate can survive before the surface
tensile stress reaches the fracture strength, and state why a thin plate
survives a thermal shock that shatters a thick one.

\textbf{Solution.} A sudden surface cooling makes the chilled face want to
contract while the still-warm interior holds it stretched, so the cooled
surface goes into tension, and glass, brittle and weak in tension, cracks from
that surface. The worst case is an instantaneous surface temperature change
$\Delta T$ against an interior that has not yet responded, for which the
surface is fully restrained in plane and the biaxial thermal stress is the
plane-stress restrained result of worked example 11,
\[
\sigma_{\text{surface}} = \frac{E\,\alpha\,\Delta T}{1-\nu}.
\]
Setting this to the fracture strength and solving for the surviving
temperature difference,
\[
\Delta T_{\text{crit}} = \frac{\sigma_f\,(1-\nu)}{E\,\alpha}
= \frac{50\times 10^6\times (1-0.22)}{70\times 10^9\times 9\times 10^{-6}}
= \frac{3.9\times 10^7}{6.3\times 10^5}
= 62\,\si{\kelvin}.
\]
A sudden surface chill of more than about $62\,\si{\kelvin}$ cracks this
glass, which matches the everyday experience that a hot glass dish tolerates
warm water but shatters in cold. The quantity in the denominator, $E\alpha$,
grouped with the strength and conductivity into the thermal-shock figure of
merit $R = \sigma_f(1-\nu)/(E\alpha)$, is why borosilicate cookware (low
$\alpha$) survives shocks that shatter soda-lime glass, and why fused silica
(very low $\alpha$) can be quenched glowing into water. The thickness enters
through the transient, not through this worst-case bound. A truly instantaneous
surface change loads the surface to the full restrained stress regardless of
thickness, but a real quench has a finite surface heat-transfer rate, and a
thin plate equalises its through-thickness temperature fast enough that the
interior follows the surface before the full $\Delta T$ builds, so the realised
stress stays below the bound. A thick plate holds its temperature gradient
longer, develops closer to the full restrained stress, and fractures at a
milder shock. The design lesson bridges worked examples 11 and 13 to the
brittle regime: a temperature field that cannot expand freely is a load set by
$E\alpha\Delta T$, and for a brittle material that load is compared not to a
yield strength but to a fracture strength that carries no ductile reserve.

\begin{mastery}
\textbf{Buckling as an eigenvalue problem.} The buckling loads scattered
through this chapter, the plate coefficient $k\pi^2D/b^2t$, the shell
external-pressure formula, the inter-ring von Mises pressure, are all special
cases of one linear-algebra statement, and seeing them that way is what lets a
finite-element code compute a buckling load for a geometry no formula covers.
Loading a structure builds two stiffness contributions. The ordinary elastic
stiffness $\mathbf{K}$ resists deformation through the material's elasticity,
exactly the assembled stiffness of the finite-element introduction. The
\engterm{geometric stiffness} $\mathbf{K}_g$ is a second, load-dependent
stiffness that captures how the membrane stresses already in the structure
help or hinder an out-of-plane deflection: a tension stiffens the structure
against buckling (a guitar string is stiffer taut), a compression softens it,
and $\mathbf{K}_g$ is proportional to the current stress state, hence to the
applied load level $\lambda$. Equilibrium of a slightly deflected shape
$\vec{v}$ under load $\lambda$ reads $(\mathbf{K} + \lambda\,\mathbf{K}_g)\,
\vec{v} = 0$. A nonzero deflection can exist at no extra load only when the
bracketed matrix is singular, which is the generalised eigenvalue problem
\[
\mathbf{K}\,\vec{v} = -\lambda\,\mathbf{K}_g\,\vec{v},
\]
whose eigenvalues $\lambda_i$ are the buckling load factors and whose
eigenvectors $\vec{v}_i$ are the buckling mode shapes. The smallest positive
$\lambda$ is the critical load, and its mode is the shape the structure buckles
into: the half-sine of the compressed plate, the diamond dimples of the axial
cylinder, the $n=6$ lobes the inter-ring vessel calculation minimised over.
Every closed-form buckling result in this chapter is the analytic eigenvalue of
this system for a geometry simple enough to solve by hand; the commercial
code's linear-buckling analysis assembles $\mathbf{K}$ and $\mathbf{K}_g$ from
the mesh and extracts the eigenvalues numerically. The method carries the same
caution the shell section raised: the eigenvalue is the perfect-structure
buckling load, so it inherits the full imperfection knock-down, and a linear
buckling eigenvalue quoted as a capacity overstates a shell's strength by the
same factor of three to five the worked examples kept subtracting. The
imperfection-sensitive cases demand the geometrically nonlinear post-buckling
analysis the von K\'arm\'an mastery box described, seeded with an imperfection
shaped like the critical eigenvector, which is why the eigenvalue analysis is
the first step of a shell-stability study and never the last.
\end{mastery}

\begin{mastery}
\textbf{Shear-deformable plate theory.} The Kirchhoff hypothesis sets the
through-thickness shear strain to zero by forcing normals to stay normal,
which is accurate while $h \le L/20$ and steadily wrong as the plate
through-thickness shear strain to zero by forcing normals to stay normal,
which is accurate while $h \le L/20$ and steadily wrong as the plate
through-thickness shear strain to zero by forcing normals to stay normal,
which is accurate while $h \le L/20$ and steadily wrong as the plate
thickens. The Mindlin-Reissner theory relaxes the hypothesis: the normal
stays straight but is allowed to rotate by an amount that need not match
the midsurface slope, so the section rotations $\phi_x, \phi_y$ become
independent fields alongside the deflection $w$. The transverse shear
strains are then $\gamma_{xz} = \phi_x + w_{,x}$ and $\gamma_{yz} = \phi_y
+ w_{,y}$, nonzero by construction, and the constitutive law adds a shear
term $Q = \kappa G h\,\gamma$ with $\kappa$ a shear-correction factor
(conventionally $5/6$ for a rectangular section) that compensates for the
assumed-uniform shear distribution. The plate now carries load by both
bending and transverse shear in series, so its compliance is the sum of a
bending compliance that scales as $1/h^3$ and a shear compliance that
scales as $1/h$. For a thin plate the bending term dominates and the theory
reduces to Kirchhoff; for a thick plate, or a sandwich panel with a
shear-soft core, the shear term is comparable or larger and the Kirchhoff
deflection is an underestimate. The three-field formulation also removes
the awkward Kirchhoff effective-shear boundary condition, because the
independent rotations supply the extra degrees of freedom an edge needs.
The pay-off runs straight into the finite-element practice of section 9.6:
Mindlin plate and shell elements need only $C^0$ continuity (the
displacement, not its slope, must be continuous across element edges),
which is why they dominate commercial shell libraries, while pure
Kirchhoff elements need the harder $C^1$ continuity. The price is shear
locking, the same artifact figure~\ref{fig:vol03ch09:shear-locking} shows
for solid elements: a thin Mindlin element with full integration starves
the bending mode and reports a stiffness far too high, and the element
libraries earn their keep with reduced integration, assumed shear strains,
or mixed formulations that switch the parasitic shear off as the plate
thins.
\end{mastery}

\begin{mastery}
\textbf{Energy methods and the Rayleigh-Ritz deflection.} The closed-form
plate solutions of this chapter cover a handful of canonical geometries and
boundary conditions; for everything else the analyst reaches for the
finite-element method or, one step before it, for the energy method that the
finite-element method discretises. The total potential energy of a bent
plate is the strain energy stored in bending minus the work done by the
load,
\[
\Pi[w] = \frac{D}{2}\!\int_A\!\left[(\nabla^2 w)^2
- 2(1-\nu)\bigl(w_{,xx}w_{,yy} - w_{,xy}^2\bigr)\right]dA
- \int_A q\,w\,dA,
\]
and the true deflection is the one that minimises $\Pi$ over all shapes that
satisfy the geometric boundary conditions. The Rayleigh-Ritz method makes
this practical: guess a deflection shape $w = \sum_i c_i\,\phi_i(x,y)$ built
from a few trial functions $\phi_i$ that each satisfy the support
conditions, substitute into $\Pi$, and minimise over the coefficients $c_i$
by setting $\partial\Pi/\partial c_i = 0$. The result is a small linear
system for the $c_i$, and even a single well-chosen trial function often
lands the centre deflection within a few percent. For a clamped square plate
under uniform load, one trial function of the form $\phi = (1-\cos 2\pi
x/a)(1-\cos 2\pi y/a)$, which vanishes with zero slope on all four edges,
recovers the centre deflection to within about three percent of the exact
value, from a single line of algebra. The method is not only a hand tool. It
is the conceptual parent of the finite-element method: replace the few
global trial functions with many local ones, one set per element, and the
Rayleigh-Ritz minimisation becomes the element stiffness assembly of the FEM
introduction. The energy method also explains why the finite-element
displacement solution is always too stiff: restricting the deflection to a
finite family of shapes can only raise the minimum of $\Pi$ above its true
value, so a displacement-based FEM under-predicts the deflection and
converges to it from below as the trial space is enriched, which is the
same one-sided convergence the mesh-convergence figure shows for the
plate-with-a-hole stress.
\end{mastery}

\section{Case study: a flat-bottom liquid storage tank}\pathtag{standard}

The two pressure-vessel cases are sized by a load that is the same at every
point of the wall: the internal or external pressure is uniform, and the
wall thickness is constant around and along the shell. A large flat-bottom
liquid storage tank breaks that symmetry. The load is hydrostatic, so it
grows linearly with depth, and the efficient wall is not constant but
stepped, thick at the base where the head is greatest and thin at the top
where it is nearly zero. We size such a tank from its service requirement
through the course-by-course shell thickness, the top wind girder that
keeps the open shell round, and the settlement and buckling notes that a
storage-tank design cannot omit, with every number carried forward. The
governing physics is hydrostatic hoop tension that varies with height, a
third distinct sizing logic alongside the air-receiver's uniform internal
pressure and the vacuum vessel's external-pressure buckling.

\textbf{The requirement.} A plant needs an atmospheric tank to hold
$2000\,\text{m}^3$ of water-like product (density $\rho = 1000\,\si{
\kilogram\per\meter\cubed}$), open to atmosphere through a vent so the only
wall load is the liquid head plus wind. We adopt a steel of yield
$\sigma_Y = 250\,\text{MPa}$ and take a code-style allowable for the design
condition of $S_d = 160\,\text{MPa}$ with a joint efficiency $E_j = 0.85$
for the vertical shell seams. A flat-bottom tank standard sets the design
point for each course a short distance above its lower seam; we use that
design-point head per course (figure~\ref{fig:vol03ch09:tank-courses}).

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-tank-courses}

\textbf{Step 1: geometry from the volume.} A flat-bottom tank is a vertical
cylinder, so the volume is $\pi R^2 H$ with $H$ the liquid height. A squat
tank is cheaper to build and easier to keep round, so choose a diameter
$D = 15\,\si{\meter}$ ($R = 7.5\,\si{\meter}$) and solve for the height,
\[
H = \frac{V}{\pi R^2} = \frac{2000}{\pi\times (7.5)^2}
= \frac{2000}{176.7} = 11.3\,\si{\meter}.
\]
Round to a fill height $H = 11.5\,\si{\meter}$ built from courses of
standard plate. Take five courses of $2.3\,\si{\meter}$ each; the geometry
of figure~\ref{fig:vol03ch09:tank-courses} is fixed.

\textbf{Step 2: shell-course thickness (hydrostatic hoop governs, course by
course).} The hoop stress at depth $d$ below the surface is the
thin-cylinder result with the local hydrostatic pressure $p(d) = \rho g d$,
\[
\sigma_\theta(d) = \frac{p(d) R}{t} = \frac{\rho g d\, R}{t}.
\]
The required thickness of a course follows from setting this to the
allowable at the course design point. The bottom course carries the
greatest head. Its design-point depth, a short way above the floor, is
about $d_1 = 10.9\,\si{\meter}$ (the fill height less roughly a foot above
the bottom seam, the one-foot rule). The required thickness is
\[
t_1 = \frac{\rho g d_1 R}{S_d E_j}
= \frac{1000\times 9.81\times 10.9\times 7.5}{160\times 10^6\times 0.85}
= \frac{8.02\times 10^5}{1.36\times 10^8}
= 5.9\times 10^{-3}\,\text{m}.
\]
Add a corrosion allowance of $1.5\,\si{\milli\meter}$ and round up to
standard plate: specify the bottom course at $t_1 = 8\,\si{\milli\meter}$.
Each course above carries a smaller design head, so the same calculation
with the reduced $d$ gives a thinner course. With the design-point depth of
each $2.3\,\si{\meter}$ course taken at its own lower-third point, the bare
required thicknesses fall roughly as $5.9, 4.7, 3.5, 2.3, 1.1\,\si{
\milli\meter}$ down the five courses; after the corrosion allowance and a
practical minimum plate of $5\,\si{\milli\meter}$ for a tank this size, the
specified courses become $8, 7, 6, 5, 5\,\si{\milli\meter}$ from bottom to
top. The wall steps down, and the upper courses are set by the handling and
corrosion minimum rather than by stress, which is the usual outcome for the
top of a tank.

\textbf{Step 3: the top wind girder.} An open-top tank, or one with only a
light roof, is a thin cylinder with a free upper edge, and a free edge of a
thin shell ovals readily under wind. Wind on a tank is not uniform external
pressure; it is a circumferential pressure distribution that pushes in on
the windward side and sucks out on the flanks, and a thin open shell
responds by going out of round, which can grow into a buckle. The fix is a
stiffening ring at or near the top, the wind girder of
figure~\ref{fig:vol03ch09:tank-courses}, sized to give the rim enough
circumferential bending stiffness to hold its shape. The required section
modulus of the girder scales with the tank diameter squared and the design
wind pressure, $Z \propto p_{\text{wind}} D^2 H_{\text{eff}}$; for a tank
this size a rolled angle or a formed-plate ring welded to the top course
supplies it. The girder also serves as a walkway curb and a fixing for the
roof, but its structural job is to keep the open shell round under wind.

\textbf{Step 4: shell buckling under wind and vacuum.} The shell between the
floor and the wind girder can buckle under the external-pressure component
of wind, the same imperfection-sensitive shell buckling as the vacuum
vessel of the previous case, now driven by wind suction rather than process
vacuum. The check is whether the unstiffened shell height between the floor
(which acts as a bottom stiffener) and the top wind girder is short enough
that the inter-stiffener buckling pressure, with its knock-down, exceeds the
wind suction. A tall tank needs one or more intermediate wind girders to
shorten the unsupported height, exactly the ring-spacing logic of the
vacuum-vessel case applied to a one-sided wind load. For our $11.5\,\si{
\meter}$ shell a single top girder is generally sufficient; a tank twice as
tall would carry an intermediate girder partway up.

\textbf{Step 5: the bottom and the settlement note.} A flat-bottom tank
rests its floor plate directly on a prepared pad, and the floor carries the
liquid weight straight into the foundation in bearing, not in bending, so
the floor plate itself is thin (a handling and corrosion minimum, not a
stress member). The structural risk at the bottom is not the floor but the
foundation: a large-diameter tank settles, and uneven settlement is the
dominant cause of storage-tank distress. Uniform settlement is benign, the
whole tank drops together; differential settlement across the diameter
tilts the shell and forces it out of round, and localised settlement under
one arc of the shell-to-bottom weld can tear that weld, the highest-stress
detail in the tank because it carries the full hydrostatic hoop tension of
the bottom course into the floor. The design specifies a compacted ring-wall
foundation under the shell, a settlement survey before and during the first
fill (the hydrotest doubles as the proof load and the settlement test), and
acceptance limits on out-of-plane and out-of-plane-tilt settlement. The
settlement note is the storage-tank counterpart of the roundness tolerance
of the vacuum vessel: a geometric tolerance that is a structural
specification, not a finishing remark.

\textbf{Verdict.} The tank is a $15\,\si{\meter}$ diameter, $11.5\,\si{
\meter}$ tall open-top steel cylinder built from five welded courses
stepping from an $8\,\si{\milli\meter}$ bottom course to a $5\,\si{
\milli\meter}$ top course, with a top wind girder sized to keep the open
shell round under wind, a thin floor plate bearing on a compacted ring-wall
foundation, and acceptance limits on differential settlement enforced
through the first-fill hydrotest. The governing check is the hydrostatic
hoop stress evaluated course by course, a calculation whose answer varies
with height where the pressure-vessel cases had a single wall; the wind
girder and the settlement limits are the two checks that the constant-wall
pressure-vessel logic does not contain. A reader who sizes this tank with a
single wall thickness set by the bottom head ships a tank that is several
times overweight at the top and still vulnerable to wind ovalling and
settlement, because the load that sets the wall is not the only load the
tank has to survive.

\section{Case study: sizing an air receiver end to end}\pathtag{standard}

The membrane formulae, the head selection, the discontinuity check, and
the buckling check are five separate calculations in the preceding
sections. A real vessel design carries one set of numbers through all of
them and ends in a verdict. We size a compressed-air receiver from its
service requirement to its specified wall and head, with every number
carried forward, and state at each step which check governs and why.

\textbf{The requirement.} A plant needs an air receiver to buffer a
reciprocating compressor: storage volume about $3\,\text{m}^3$, maximum
allowable working pressure $p = 1.6\,\text{MPa}$ gauge (a common
$16\,\text{bar}$ rating), service temperature near ambient, working
fluid dry compressed air. The vessel is a horizontal cylinder closed by
two heads, with a single top nozzle for the relief valve and outlet
(figure~\ref{fig:vol03ch09:vessel-layout}). The material is a plain
carbon-steel plate of yield $\sigma_Y = 250\,\text{MPa}$ and a
code-style allowable stress $S = 138\,\text{MPa}$ (a representative
value for this class of steel at modest temperature; a real design
reads $S$ from the material tables of the governing code). We adopt a
joint efficiency $E_j = 0.85$, the value for a spot-radiographed
double-welded butt seam.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-vessel-layout}

\textbf{Step 1: geometry from the volume.} Choose an inner diameter
$D_i = 1.2\,\si{\meter}$, so $R = 0.6\,\si{\meter}$. The cylindrical
shell of tangent length $L$ holds $\pi R^2 L$; the two hemispherical-
equivalent heads add roughly $\tfrac{4}{3}\pi R^3$ between them. Setting
the total to $3\,\text{m}^3$,
\[
\pi (0.6)^2 L + \tfrac{4}{3}\pi (0.6)^3 = 3
\;\Rightarrow\;
1.131\,L + 0.905 = 3
\;\Rightarrow\;
L = 1.85\,\si{\meter}.
\]
Round the tangent length up to $L = 3.0\,\si{\meter}$ to leave margin
and to suit standard plate, giving a generous $\sim 4.3\,\text{m}^3$;
the geometry of figure~\ref{fig:vol03ch09:vessel-layout} is fixed.

\textbf{Step 2: cylindrical shell thickness (hoop governs).} The hoop
stress is the governing membrane stress. Sizing on the code-style
relation $t = pR/(S E_j)$, which is the membrane hoop formula with the
allowable reduced by the joint efficiency,
\[
t_{\text{shell}} = \frac{p R}{S E_j}
= \frac{1.6\times 10^6\times 0.6}{138\times 10^6\times 0.85}
= \frac{9.6\times 10^5}{1.173\times 10^8}
= 8.19\times 10^{-3}\,\text{m}
= 8.2\,\si{\milli\meter}.
\]
Add a corrosion allowance of $1.5\,\si{\milli\meter}$ and round to a
standard plate: specify $t_{\text{shell}} = 10\,\si{\milli\meter}$. The
as-specified hoop stress on the corroded wall ($8.5\,\si{\milli\meter}$
after allowance) is $\sigma_\theta = pR/t = 1.6\times 10^6\times 0.6/
(8.5\times 10^{-3}) = 113\,\text{MPa}$, comfortably under the
$138\,\text{MPa}$ allowable.

\textbf{Step 3: head selection and thickness.} A hemispherical head
carries $pR/2t$ and would size at $t = pR/(2 S E_j) = 4.1\,\si{
\milli\meter}$, half the shell. A hemisphere is deep and costly to
form, though, so we choose a $2{:}1$ ellipsoidal head, the common
compromise. The ellipsoidal head's governing thickness, from the code
relation $t = pD_i/(2 S E_j - 0.2 p)$, is
\[
t_{\text{head}} = \frac{1.6\times 10^6\times 1.2}
{2\times 138\times 10^6\times 0.85 - 0.2\times 1.6\times 10^6}
= \frac{1.92\times 10^6}{2.346\times 10^8 - 3.2\times 10^5}
= 8.2\times 10^{-3}\,\text{m},
\]
the same as the shell thickness to two figures. Specify the head at the same
$10\,\si{\milli\meter}$ as the shell, which also eases the welding of
two equal-thickness edges. The thickness ordering across head types is
the one shown in figure~\ref{fig:vol03ch09:head-types}: the
hemispherical head is thinnest, the ellipsoidal and torispherical heads
sit near the shell thickness, and a flat head would be several times
thicker because it carries the pressure in bending.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-head-types}

\textbf{Step 4: head-to-shell junction.} The ellipsoidal head and the
cylinder grow by slightly different radial amounts under pressure, the
discontinuity-stress mechanism of worked example 8. With $R = 0.6\,
\si{\meter}$ and $t = 10\,\si{\milli\meter}$, the boundary-layer width
is of order $\sqrt{R t} = \sqrt{0.6\times 0.01} = 0.077\,\si{\meter}$,
so the local bending is confined to a band about $80\,\si{\milli\meter}$
from the junction. We detail the junction with the head's own knuckle
radius (the $2{:}1$ ellipsoidal profile supplies it) and avoid any
abrupt thickness step, keeping the two welded edges at equal thickness.
No extra wall is needed here for a vessel of this modest pressure; the
detail is the fix.

\textbf{Step 5: nozzle reinforcement.} The top nozzle cuts a hole in
the shell, and a hole removes load-carrying metal, exactly the
stress-concentration mechanism of worked example 4. The area-
replacement rule of pressure-vessel codes requires that the metal
removed by the opening be replaced by reinforcement within a defined
zone around the nozzle. For a nozzle of inner diameter $d = 100\,
\si{\milli\meter}$, the area removed is $A_{\text{rem}} = d\,t_{\text{
required}} = 0.100\times 8.2\times 10^{-3} = 8.2\times 10^{-4}\,
\text{m}^2$ (using the required, not the specified, wall). The excess
wall already present (the specified $10\,\si{\milli\meter}$ against the
$8.2\,\si{\milli\meter}$ required, over the reinforcement zone) supplies
part of this; the nozzle neck supplies more; a reinforcing pad makes up
any shortfall. We note the check rather than work the full area
balance: the design intent is that the opening not lower the vessel's
pressure capacity, achieved by replacing the removed area locally.

\textbf{Step 6: external-pressure and buckling check.} An air receiver
is an internal-pressure vessel, but it can see external pressure during
a vacuum event (a blocked inlet during pump-out, or steam condensation
if it were ever steamed and sealed). The relevant failure is inward
buckling, not yield. The long-cylinder ring-buckling pressure is
$p_{\text{cr}} = E t^3/[4(1-\nu^2)R^3]$; with $E = 200\,\text{GPa}$,
$t = 8.5\times 10^{-3}\,\si{\meter}$ (corroded), $R = 0.6\,\si{\meter}$,
\[
p_{\text{cr}} = \frac{200\times 10^9\times (8.5\times 10^{-3})^3}
{4\times 0.91\times (0.6)^3}
= \frac{200\times 10^9\times 6.14\times 10^{-7}}{0.786}
= 1.56\times 10^5\,\text{Pa}.
\]
A $0.3$ imperfection knock-down brings this to about $47\,\text{kPa}$,
below a full atmosphere of external differential ($101\,\text{kPa}$).
The bare shell would not survive a hard vacuum. The verdict at this
step is to fit a vacuum-breaker (anti-vacuum valve) sized so the
credible external differential cannot exceed the knocked-down capacity,
the standard and cheap fix; ring stiffeners would be the alternative if
a vacuum-rated vessel were required.

\textbf{Verdict.} The vessel is a $1.2\,\si{\meter}$ inner-diameter,
$3.0\,\si{\meter}$ tangent-length carbon-steel cylinder, $10\,\si{
\milli\meter}$ wall, with two $2{:}1$ ellipsoidal heads also at
$10\,\si{\milli\meter}$, a top nozzle reinforced by the local excess
wall plus a pad as the area balance requires, equal-thickness welds at
the head-to-shell junctions, full radiography downgraded to spot with
$E_j = 0.85$, a $1.5\,\si{\milli\meter}$ corrosion allowance, and a
vacuum-breaker against external-pressure collapse. The governing check
is the cylindrical hoop stress; the head, junction, and nozzle checks
each refine the detail without changing the wall; the external-pressure
check changes nothing structural but adds a protective device. The
membrane sizing arithmetic and the head and external-pressure relations
are implemented in \repopath{code/pressure_vessel.py}. A reader who
takes the membrane hoop result as the whole design ships a vessel that
collapses on the first vacuum event; the discipline is to carry the
numbers through every check, not only the one that sizes the wall.

\section{Case study: a ring-stiffened vessel under external
pressure}\pathtag{standard}

The air-receiver case is sized by yield: the internal pressure pushes the
wall into tension, and the wall thickness follows from the hoop stress.
Reverse the load and the governing physics changes completely. A vessel
under external pressure fails by inward buckling, a stability problem,
long before any stress approaches yield, and the design is set by the
shell's geometric stiffness rather than its strength. We size a
horizontal cylindrical vacuum vessel from its service requirement through
the bare-shell buckling check (which it fails), the ring-stiffening that
rescues it, the ring sizing, and the manufacturing tolerance that the
knock-down factor encodes, with every number carried forward to a
verdict. This is the external-pressure counterpart of the air-receiver
case and shares none of its governing logic.

\textbf{The requirement.} A process plant needs a horizontal cylindrical
vessel to run under vacuum, with the wall also able to take a credible
upset of partial external overpressure during a clean-in-place cycle. We
adopt a design external differential of $p = 0.15\,\text{MPa}$ gauge,
which covers a full atmosphere of vacuum ($0.101\,\text{MPa}$) with a
margin of about $1.5$ on the loading itself, before any structural
factor. The vessel is a carbon-steel cylinder, inner diameter $D = 1.5\,
\si{\meter}$ ($R = 0.75\,\si{\meter}$), tangent length $6\,\si{\meter}$,
closed by dished ends, with the material yield $\sigma_Y = 250\,\text{
MPa}$ and $E = 200\,\text{GPa}$, $\nu = 0.3$. The internal process does
not pressurise the vessel above atmospheric, so there is no hoop-tension
sizing case at all; the wall exists to resist collapse
(figure~\ref{fig:vol03ch09:ring-stiffened-vessel}).

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-ring-stiffened-vessel}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-buckling-vs-spacing}

\textbf{Step 1: try the bare shell.} Pick a thin wall, $t = 6\,\si{
\milli\meter}$, and ask whether the bare cylinder survives the external
pressure as a long cylinder. The long-cylinder ring-collapse pressure is
\[
p_{\text{cr}} = \frac{E t^3}{4(1-\nu^2) R^3}
= \frac{200\times 10^9 \times (0.006)^3}
{4\times 0.91\times (0.75)^3}
= \frac{200\times 10^9 \times 2.16\times 10^{-7}}{1.536}
= 2.81\times 10^4\,\text{Pa}.
\]
The bare-shell critical pressure is $28\,\text{kPa}$, and a $0.8$
imperfection knock-down (more generous than the axial-compression case,
because external-pressure buckling is less imperfection-sensitive) brings
it to about $22\,\text{kPa}$. This is well below a single atmosphere of
vacuum, let alone the $150\,\text{kPa}$ design differential. The bare
$6\,\si{\milli\meter}$ cylinder collapses on the first pump-down. We
could thicken the wall, but the long-cylinder capacity scales as $t^3$,
so reaching $150\,\text{kPa}$ with the bare shell would need
$t \approx 6\times (150/22)^{1/3} \approx 11\,\si{\milli\meter}$ working
against the knocked-down value, and a thicker shell still buckles as a
long cylinder. The efficient move is not more wall, it is to shorten the
unsupported length.

\textbf{Step 2: add ring stiffeners.} A ring stiffener forces a node in
the buckling mode, so the shell buckles between rings as a short cylinder
rather than over its full length as a long one. The short-cylinder
buckling pressure rises steeply as the ring spacing $L_s$ falls
(figure~\ref{fig:vol03ch09:buckling-vs-spacing}); the relevant relation
is the von Mises inter-ring formula, which minimises over the number of
circumferential buckling lobes $n$,
\[
p_{\text{cr}} = \frac{E t}{R}\left[
\frac{(\pi R/L_s)^4}{\bigl(n^2 - 1 + \tfrac{1}{2}(\pi R/L_s)^2\bigr)
\bigl(n^2 + (\pi R/L_s)^2\bigr)^2}
+ \frac{t^2}{12(1-\nu^2) R^2}\,(n^2 - 1)
\right].
\]
Choosing a ring spacing $L_s = 1.5\,\si{\meter}$, so $L_s/R = 2.0$, and
minimising over $n$ gives the worst mode at $n = 6$ lobes and an
inter-ring critical pressure of about $510\,\text{kPa}$. Applying the
$0.8$ external-pressure knock-down leaves a working capacity near
$410\,\text{kPa}$, a margin of about $2.7$ over the $150\,\text{kPa}$
design differential. The $6\,\si{\milli\meter}$ wall is now adequate
between rings, and the design splits the $6\,\si{\meter}$ tangent length
into four bays with three internal rings (plus the end closures, which
act as terminal stiffeners). The thickness-cubed penalty that made the
bare shell hopeless is sidestepped by the ring spacing, which enters the
short-cylinder capacity far more strongly than the wall thickness does.

\textbf{Step 3: size the rings.} The shell sheds its inward pressure to
the rings as a radial line load over each ring's tributary width
$L_s$, and each ring must carry that load without itself collapsing as a
ring. The ring's own buckling (the \engterm{general-instability} mode,
in which rings and shell deflect together into a few large lobes) is held
off by giving each ring enough flexural stiffness. The standard sizing
requires a minimum ring second moment of area; with a design factor of
$3$ on the ring against the tributary external load,
\[
I_{\text{ring}} \ge \frac{p\,L_s\,R^3\,\text{FS}}{3 E}
= \frac{0.15\times 10^6 \times 1.5\times (0.75)^3\times 3}
{3\times 200\times 10^9}
= 4.7\times 10^{-7}\,\text{m}^4 = 47\,\si{\centi\meter}^4.
\]
A flat-bar ring $90\,\si{\milli\meter}$ deep by $10\,\si{\milli\meter}$
thick has a second moment of $b d^3/12 = 0.010\times (0.090)^3/12 =
61\,\si{\centi\meter}^4$ about its own centroid, already above the
requirement. Crediting the effective shell strip that acts with the ring
(a width of about $1.56\sqrt{R t}$ on each side, here a total of about
$210\,\si{\milli\meter}$ of shell) raises the combined section's second
moment to roughly $180\,\si{\centi\meter}^4$, several times the
requirement, which is the usual outcome: once the ring is deep enough to
matter at all, the shell strip it carries makes it generously stiff.
Specify the three rings as $90\times 10\,\si{\milli\meter}$ flat bars,
welded internally with a continuous fillet to the shell.

\textbf{Step 4: the knock-down is a tolerance.} The $0.8$ factor applied
in steps 1 and 2 is not a fudge; it is the imperfection sensitivity of
the section made into a number, and it is the design's statement about
manufacturing tolerance. An out-of-round shell buckles at a lower
pressure than a perfect one, and external-pressure codes tie the
permitted out-of-roundness directly to the buckling capacity: the
maximum difference between the largest and smallest measured diameters is
held to a small fraction of the radius (a common limit is about $1\,\%$
of the diameter). The fabrication drawing therefore carries a roundness
tolerance, and the as-built vessel is measured against it before service.
A shell built to a looser roundness than the design assumed carries a
knock-down worse than $0.8$, and its true capacity can fall below the
$150\,\text{kPa}$ requirement even though every dimension on the drawing
was nominally met. The tolerance is the structural specification, not a
finishing note.

\textbf{Verdict.} The vessel is a $1.5\,\si{\meter}$ inner-diameter,
$6\,\si{\meter}$ tangent-length carbon-steel cylinder with a $6\,\si{
\milli\meter}$ wall, stiffened by three internal $90\times 10\,\si{
\milli\meter}$ ring frames at $1.5\,\si{\meter}$ spacing, with dished
ends acting as terminal stiffeners, built to a $1\,\%$ out-of-roundness
tolerance, and protected by the design margin of about $2.7$ on the
knocked-down inter-ring buckling capacity against the $150\,\text{kPa}$
external differential. The governing check is the inter-ring shell
buckling pressure, a stability calculation in which the ring spacing is
the dominant parameter and the wall thickness is secondary; the ring
sizing and the roundness tolerance are the two checks that keep the
stiffened design from failing in the modes the bare-shell calculation
cannot see. A reader who sizes this vessel by the internal-pressure hoop
formula of the air-receiver case computes a wall thickness of nearly
zero, because there is no internal pressure, and ships a vessel that
implodes on the first pump-down. The external-pressure problem is a
different problem, and the discipline is to recognise which problem is in
front of you before reaching for a formula.

\section{A spherical gas-storage vessel}\pathtag{standard}

The three case studies size cylinders. A short fourth example sizes the
geometry the membrane analysis singles out as most efficient, the sphere,
and shows what the efficiency costs elsewhere. A spherical gas holder, the
Horton sphere of figure~\ref{fig:vol03ch09:storage-sphere}, stores a gas
at pressure in the shape that carries internal pressure at the lowest
wall stress.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-storage-sphere}

\textbf{The requirement and the wall.} A plant stores
$V = 1000\,\text{m}^3$ of gas at a working pressure $p = 0.8\,\text{MPa}$
gauge in a sphere, with steel of allowable stress $S = 138\,\text{MPa}$
and joint efficiency $E_j = 0.85$. The radius follows from the volume,
$R = (3V/4\pi)^{1/3} = (3\times 1000/4\pi)^{1/3} = (238.7)^{1/3} =
6.2\,\si{\meter}$. The membrane stress is equal in every direction,
$\sigma = pR/2t$, so the wall thickness from the code-style relation
$t = pR/(2 S E_j)$ is
\[
t = \frac{0.8\times 10^6 \times 6.2}{2\times 138\times 10^6\times 0.85}
= \frac{4.96\times 10^6}{2.35\times 10^8}
= 2.1\times 10^{-2}\,\text{m} = 21\,\si{\milli\meter}.
\]
Add a corrosion allowance and round to standard plate: specify
$t = 24\,\si{\milli\meter}$. A cylinder of the same radius and pressure
would need twice this wall on its hoop, $t_{\text{cyl}} = pR/(S E_j) =
42\,\si{\milli\meter}$, so the sphere saves roughly half the shell steel.
For a vessel this large the saving is many tonnes of plate, which is why
high-volume high-pressure storage is built spherical despite the cost of
forming and welding the doubly curved petal plates.

\textbf{Where the efficiency is paid back.} The sphere has no membrane
weakness, so the design moves straight to the discontinuities. The sphere
rests on legs or columns, and each support introduces a local radial and
meridional force into a shell that carries pure membrane tension
everywhere else. The support reaction sets up a bending boundary layer in
the shell over a band of order $\sqrt{R t} = \sqrt{6.2\times 0.024} =
0.39\,\si{\meter}$ around each support attachment, and the local bending
stress there can rival the membrane stress. The fix is a wider support
saddle or a reinforcing pad that spreads the reaction over a larger arc,
the same discontinuity-stress logic as the cylinder-to-head junction,
applied now to a column-to-shell junction. The nozzles, the manway, and
the support skirt are each a hole or a load point in an otherwise uniform
membrane, each needing local reinforcement. The sphere proves the
chapter's recurring lesson in its cleanest form: the membrane formula
sizes the wall in one line, and every detail that actually governs the
design lives at a discontinuity the membrane formula cannot see.

\section{Vignette: a micromachined pressure sensor}\pathtag{standard}

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-mems-diaphragm}

The plate theory of this chapter is usually met at the scale of a floor
slab or a vessel wall, but the same equations govern a silicon membrane a
few micrometres thick, and the micromachined pressure sensor is where the
clamped-plate relations earn their keep in a device sold by the billion.
A micro-electromechanical (MEMS) pressure sensor etches a thin square
diaphragm into a silicon wafer over a sealed reference cavity
(figure~\ref{fig:vol03ch09:mems-diaphragm}). The pressure to be measured
bows the diaphragm into the cavity, and the device reads the bending it
produces.

Take a silicon diaphragm of side $2a = 1\,\si{\milli\meter}$ and thickness
$h = 20\,\si{\micro\meter}$, clamped on all four edges, with silicon's
$E = 170\,\text{GPa}$ and $\nu = 0.22$. Its flexural rigidity is $D =
E h^3/[12(1-\nu^2)] = 170\times 10^9\times (20\times 10^{-6})^3/[12\times
0.952] = 1.19\times 10^{-4}\,\text{N}\cdot\text{m}$. Under a full-scale
pressure $p = 100\,\text{kPa}$ the clamped square plate deflects at its
centre by about $w \approx 0.0138\,p\,(2a)^4/(D\cdot 16)$; carrying the
clamped-square coefficient through, the centre deflection is of order
$1\,\si{\micro\meter}$, one twentieth of the diaphragm thickness, which
places the device squarely inside the small-deflection linear range where
the response is linear in pressure.

The transduction reads strain, not deflection. The bending strain in a
clamped plate is largest at the midpoints of the clamped edges, where the
plate hogs over its built-in support, so the sensor diffuses piezoresistors
at those four points and wires them as a Wheatstone bridge. Two gauges sit
where the edge bends the diaphragm one way and two where it bends the other,
so the bridge differences the strains and doubles the signal while
cancelling the temperature drift common to all four. The output is a voltage
linear in the pressure over the small-deflection range, and the plate
theory sets both the sensitivity (through the edge bending strain per unit
pressure) and the upper limit of the usable range (the pressure at which the
centre deflection approaches the diaphragm thickness and the von K\'arm\'an
membrane stretching turns the response nonlinear). The burst pressure is
set by the peak edge stress reaching the fracture strength of silicon, which
is brittle and strong in compression but sensitive to surface flaws in
tension, so the diaphragm is oriented and passivated to keep the tensile
edge stress off the weakest crystal planes. The whole device is a clamped
plate whose deflection is read as a strain, and every number in its
datasheet traces back to the clamped-plate relations of the plate section
of this chapter, applied at a scale where the plate is smaller than a grain
of rice.

\section{Vignette: the shell of a wind-turbine tower}\pathtag{standard}

A large wind-turbine tower is a tapered steel tube a hundred metres tall
that carries the rotor and nacelle at its top, and it is a thin cylindrical
shell whose design is governed by the shell-buckling and fatigue mechanisms
of this chapter rather than by simple bending strength. A representative
tower has a base diameter of about $4\,\si{\meter}$, a wall thickness that
steps down from roughly $30\,\si{\milli\meter}$ at the base to $12\,\si{
\milli\meter}$ near the top, and a nacelle-plus-rotor mass of order $300\,
\text{tonnes}$ perched at the top. The dominant load is not the weight but
the thrust of the wind on the rotor, which applies a large bending moment at
the base, and the base of the tower is therefore a thin cylinder in bending.

A thin cylinder in bending does not fail the way a solid beam does. The
compression side of the bent tube carries a membrane compression that can
buckle the shell locally into diamond-shaped dimples long before the
material yields, the axially compressed cylinder that worked example 17
showed is the most imperfection-sensitive shell mode of all. The base
diameter-to-thickness ratio here is $D/t = 4/0.030 = 133$, thin enough that
the local buckling stress with its severe knock-down, not the yield stress,
sets the wall thickness on the compression side. The tower is also prone to
the Brazier ovalisation the shell section named: the bending flattens the
circular section, which lowers the effective bending stiffness and feeds the
local buckling. The design carries the wall thickness needed to keep the
knocked-down local buckling stress above the peak compressive membrane
stress at the base, and it is that stability check, not a stress-versus-yield
comparison, that governs.

Two further mechanisms make the tower a shell problem rather than a beam
problem. The tower is a slender structure with a heavy top mass, so it has a
low fundamental bending frequency, typically a few tenths of a hertz, and
that frequency must be kept clear of the rotor's once-per-revolution and
blade-passing excitation frequencies, or the tower resonates and fatigues,
the plate-frequency lesson of worked example 14 at the scale of a
skyscraper. The whole tower sees millions of load cycles over its life as
the wind gusts and the rotor turns, so every welded seam and every door or
duct opening in the shell is a fatigue detail, and the flange bolts that
join the tower sections are sized against fatigue, not against static
strength. The door opening at the base, a hole cut in the most heavily
loaded region of the shell, is reinforced exactly as a vessel nozzle is,
by the area-replacement logic of worked example 28, because a
square-cornered opening in a fatigue-loaded shell is the Comet cabin cutout
in another guise. The tower gathers nearly every mechanism of this chapter,
shell buckling, imperfection sensitivity, ovalisation, resonance, fatigue at
stress concentrations, into a single structure, which is why a wind-turbine
tower is designed by shell analysis and finite-element models rather than by
the beam formulas its slender shape might suggest.

\section{Vignette: the composite-overwrapped pressure vessel}\pathtag{standard}

The pressure-vessel worked examples of this chapter size an isotropic metal
wall from a single membrane stress. A composite-overwrapped pressure vessel
(COPV), the tank that stores breathing air on a firefighter's back, hydrogen
at $700\,\text{bar}$ in a fuel-cell car, and helium on a launch vehicle,
carries its pressure in a wound fibre shell whose stiffness is direction
dependent, and it is where the anisotropy section of this chapter meets the
shell section in a device sold by the million. The construction is a thin
inner liner, a metal or polymer barrier that holds the gas but carries little
load, wrapped in many layers of carbon or glass fibre set in resin, each layer
laid at a chosen angle. The fibres carry the load; the liner only seals.

The winding angle is the whole design, and the reason is the two-to-one
stress ratio of the thin cylinder. The hoop stress is twice the axial stress,
so a wall that carried the same strength in every direction, as an isotropic
metal does, is twice as strong as it needs to be in the axial direction and
wastes half its axial material. A wound composite escapes the waste by putting
fibre where the load is. \engterm{Hoop windings}, laid nearly circumferentially
at close to $90^\circ$ to the axis, carry the hoop stress; \engterm{helical
windings}, laid at a shallower angle, carry the axial stress and, because they
cross the dome ends, tie the two heads together. The classic result of
netting analysis, the fibre-only membrane balance, is that a geodesic helical
winding angle of about $54.7^\circ$ (the angle whose tangent squared is two)
balances a vessel wound with helical layers alone against the two-to-one hoop
and axial stresses, which is why $54.7^\circ$ recurs across wound-vessel
practice. Real vessels combine a majority of hoop layers over the cylinder
with helical layers that also form the domes, and the layup is the anisotropic
design decision the classical-laminate-theory discussion named: the stacking
sequence sets not only the strength but which direction carries which load.

The failure behaviour is where the composite vessel departs most sharply from
the metal one, and it departs in a direction that favours the composite. A
metal vessel that is over-pressurised yields and then bursts, and the burst is
a fragmenting release of the stored gas energy. A well-designed COPV instead
leaks before it bursts: the liner cracks and vents the gas through the porous
overwrap before the fibre shell reaches its ultimate strength, a graceful
failure the certification standards require and test for. The vessels are also
proof-tested and, in the most demanding aerospace use, monitored by acoustic
emission during the proof cycle, listening for the sound of individual fibre
breaks as an early sign of a weak wall, the validation discipline of this
chapter's VVUQ section applied by ear. The stress-rupture behaviour of the
fibre, its slow loss of strength under sustained load over years, is the life
limit that a metal vessel does not share, and it is why a COPV carries a
service-life stamp that a steel cylinder does not. The device gathers the
chapter's threads into one wall: the membrane stress ratio of the thin
cylinder sets the winding angles, the anisotropic constitutive law of the
overwrap sets the layup, the discontinuity discipline governs the dome and the
boss where the fibres turn and the valve threads in, and the
leak-before-burst requirement is a failure-mode specification carried in the
design from the first layer.

\section{Failure: when finite-element analyses lie}\pathtag{core}

A finite-element analysis is a numerical method; its output is a
set of numbers. The numbers are correct only if the model
(geometry, material properties, boundary conditions, load
distribution), the discretisation (mesh, element type), the solver
(numerical parameters, convergence tolerances), and the
interpretation (which output quantity, at which point, in which
unit) are all correct. The failure section of this chapter names
the classes of error that produce wrong FEM results and the
diagnostics that catch them.

\subsection{Modelling errors}

The simulation solves the wrong problem.

\begin{itemize}[leftmargin=*]
  \item \engterm{Wrong material model.} An anisotropic material is
    analysed as isotropic; a nonlinear material is analysed as
    linear; a temperature-dependent material is analysed at the
    wrong temperature; the elastic constants are taken from a
    handbook for the wrong alloy. The diagnostic is a coupon test
    on the actual material under the actual conditions.
  \item \engterm{Wrong boundary conditions.} A bolted joint is
    modelled as a rigid weld; a contact surface is modelled as
    bonded; a frictional interface is modelled as frictionless. A
    structure that is hinged in reality but modelled as
    moment-resisting will be predicted as stiffer and stronger
    than the real structure.
  \item \engterm{Wrong load distribution.} A pressure load is
    applied uniformly when the real pressure has a gradient; a
    point load is applied at a node when the real load is spread
    over a contact area; a dynamic load is analysed statically
    when inertia or rate effects matter. The diagnostic is a
    careful free-body diagram (chapter 1) of the system, applied
    before the finite-element analysis is built.
  \item \engterm{Wrong geometry.} The CAD model differs from the
    as-built geometry (chamfers omitted, fillets simplified,
    tolerances not represented). The diagnostic is to compare the
    CAD-to-mesh transcription against the manufacturing drawing
    and against the as-built measurements.
\end{itemize}

\subsection{Discretisation errors}

The simulation solves the right problem incorrectly.

\begin{itemize}[leftmargin=*]
  \item \engterm{Coarse mesh at a stress concentration.} The peak
    stress is under-predicted; the predicted location of failure
    moves away from the actual location. The diagnostic is mesh
    refinement and a mesh-convergence study.
  \item \engterm{Linear elements on a bending-dominated thin
    structure.} Shear locking causes linear tetrahedra and
    hexahedra to over-predict the stiffness of thin plates and
    shells. The diagnostic is to use shell elements, reduced-
    integration solid elements, or quadratic solid elements.
  \item \engterm{Volumetric locking in nearly incompressible
    material.} Standard displacement elements over-predict the
    stiffness of nearly incompressible materials. The diagnostic
    is to use mixed displacement-pressure elements.
  \item \engterm{Poor element quality at automatic mesh
    refinement.} The mesh generator produces thin sliver
    elements at curved boundaries or near point loads. The
    diagnostic is element-quality checks (aspect ratio,
    skewness, warpage) on every mesh.
\end{itemize}

\subsection{Interpretation errors}

The simulation produces correct numbers that the engineer reads
incorrectly.

\begin{itemize}[leftmargin=*]
  \item \engterm{Reading a stress value at a node where it is
    inaccurate.} Element stresses are typically most accurate at
    the Gauss points, less accurate at the element edges, and
    least accurate at the nodes (where they are computed by
    averaging or extrapolation from adjacent elements). A node
    stress at a stress concentration, in a coarse mesh, can be
    badly inaccurate.
  \item \engterm{Failing to recognise a singularity}. A
    re-entrant corner produces an unbounded stress in the
    elastic-theory limit; the finite-element peak stress at the
    corner increases without bound as the mesh is refined.
    Interpreting the corner stress as a failure criterion gives a
    progressively more conservative design as the mesh is refined,
    which is the opposite of convergence to a sensible answer. The
    engineer must recognise the singularity (the corner is a
    geometric feature with no physical analogue at sufficient
    resolution) and apply a different failure criterion (J-integral,
    notch-stress, structural-stress methods).
  \item \engterm{Unit confusion}. Forces in newtons, stresses in
    pascals, lengths in millimetres; or forces in pounds, stresses
    in psi, lengths in inches. A unit confusion gives results
    wrong by factors of $10^3$ or $10^6$. The diagnostic is to
    check units on every input and every output, and to verify
    that the units of the result are the units the engineer
    expected.
  \item \engterm{Wrong factor of safety}. The finite-element peak
    stress is taken as the design stress, without considering the
    stress concentration's locality, the material's fatigue
    properties, or the load uncertainty. The diagnostic is the
    full code-based design check (chapter 8) applied to the FEM
    result.
\end{itemize}

\subsection{The thin-mirror analogue: Hubble's metrology chain}

The Hubble Space Telescope primary mirror, fabricated by
Perkin-Elmer between 1979 and 1981, is the canonical case study
of a precision-fabrication failure in which the test instrument
itself was wrong \cite{acc:nasa-hubble-optical-systems-1990}.
The mirror was figured to the wrong shape because the reflective
null corrector used to measure the figure during polishing was
misassembled: a field lens was mispositioned by about $1.3\,
\si{\milli\meter}$ along the optical axis, the result of a
misinterpretation of a measurement target on a metering rod
during assembly. The error in the test instrument propagated into
the polished mirror's figure: the outer edge was about $2\,\si{
\micro\meter}$ too flat, an error large enough to ruin
diffraction-limited imaging.

The Hubble case is not a finite-element failure; the mirror was
fabricated mechanically and measured optically, with no
finite-element analysis in the verification chain. Its relevance
to the present chapter is the parallel discipline: a numerical
result is only as good as the verification chain that established
its trustworthiness. The reflective null corrector was treated as
the authoritative figure measurement, with no independent
verification of the test instrument itself; two auxiliary optical
tests showed discrepancies and were set aside. The engineering
discipline that should have caught the error was the
\engterm{independent verification of the test instrument} (in FEM
language: code verification of the simulation). The discipline
that should have caught the residual error was the
\engterm{anomaly resolution} that takes seriously every
discrepancy between an authoritative result and an auxiliary
check (in FEM language: validation against an independent
analysis or test).

The thin-mirror connection to plate and shell mechanics is also
direct. The Hubble primary is a $2.4 \,\text{m}$ diameter ULE
glass shell with a mass of about $830 \,\si{\kilogram}$ and an
average thickness of about $300\,\si{\milli\meter}$; in
ground-based optics the figure error from gravity sag and
mounting forces is computed by FEM analysis of the shell as a
thin curved plate. Modern lightweight space telescopes use even
thinner mirrors (the Herschel primary was $3.5 \,\text{m}$
diameter and only $\sim 20\,\si{\milli\meter}$ thick), and the
mechanical analysis of figure error under thermal and gravity
load is wholly an FEM exercise. The discipline that catches
errors in those analyses is the same VVUQ discipline named in
section 9.7.

\subsection{The square-corner analogue: the Comet cabin cutouts}

The stress-concentration discipline of this chapter has a structural
counterpart as sharp as the Hubble metrology case. The de Havilland
Comet, the first jet-powered passenger airliner, suffered two in-flight
cabin disintegrations in 1954, traced by the Royal Aircraft
Establishment to fatigue cracks that initiated at the square-cornered
cutouts in the pressurised cabin skin and propagated under repeated
pressurisation cycles \cite{acc:cohen-comet-report-1955}. The cabin
was a thin cylindrical pressure vessel; each flight took it through one
full pressure cycle of about $0.55\,\si{\bar}$ differential. The
cutouts had small fillet radii, and a small fillet radius at the edge
of a cutout in a loaded plate produces exactly the stress
concentration that figure~\ref{fig:vol03ch09:kirsch-hole} quantifies
for a circular hole, except that the square corner concentrates the
stress more sharply than a round hole does. The local peak exceeded
the material's fatigue limit; a crack initiated, grew under cycling,
reached its critical length, and ran.

The connection to FEM practice is the singularity error of section
9.7. A re-entrant or sharp corner produces an unbounded stress in the
elastic limit; an FEM model of the Comet cutout meshed coarsely would
have reported a finite, reassuring peak that rose without bound as the
mesh refined. The discipline that catches it is the one named earlier:
recognise the geometric singularity, refine the corner to a real
fillet radius, and apply a fatigue criterion rather than a static
peak-stress check. The post-Comet response, full-airframe fatigue
testing and the generous radiusing of every cutout, is the structural
analogue of the mesh-convergence and singularity discipline of this
chapter. A round window is a stress-concentration fix you can see from
your seat.

\begin{principle}[A FEM result is a hypothesis, not a fact]
A finite-element output is a numerical prediction of how the
modelled system would respond to the modelled load. The
prediction is a fact about the model; it is a hypothesis about
the physical system. The discipline that turns the hypothesis
into a fact about the physical system is verification (the
solver solves the model correctly), validation (the model
captures the physics), and uncertainty quantification (the
inputs propagate to bounded outputs). A FEM output without
verification, validation, and uncertainty quantification is a
colourful screenshot, not an engineering claim.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Analytical plus FEM, with reconciliation]
The reader solves a two-dimensional elasticity problem by hand,
solves the same problem with a finite-element package, compares
the two results, identifies the source of any disagreement, and
reports the reconciliation. The deliverable is a thirty- to
fifty-page engineering report.

\textbf{Track}. Analytical plus simulation. No physical build.
\textbf{Hazard class}: none.

\textbf{Problem}. A flat steel plate, $200\,\si{\milli\meter}
\times 100 \,\si{\milli\meter}$, $5\,\si{\milli\meter}$ thick, with
a central circular hole of $20\,\si{\milli\meter}$ diameter, is
loaded uniaxially in the long direction with a uniform tension of
$50\,\text{MPa}$ applied to its short edges. The long edges and
the hole boundary are traction-free. Find the maximum stress at
the hole.

\input{volumes/03-force/ch09-plates-shells-solid-mechanics/figures/fig-kirsch-hole}

\textbf{Analytical solution}. The Kirsch (1898) solution for an
infinite plate with a circular hole under remote uniaxial tension
gives a stress concentration factor of $3.0$ at the hole
(figure~\ref{fig:vol03ch09:kirsch-hole}). With the
applied stress of $50\,\text{MPa}$, the peak hoop stress at the
hole is $150\,\text{MPa}$. The reader writes out the Kirsch
solution from first principles (chapter 3 of this volume, section
on plane stress), verifies the stress concentration factor, and
notes the role of the plate's finite width in modifying the
analytical result (the finite-width correction for a $D/W = 0.2$
ratio raises the stress concentration factor to approximately
$3.13$).

\textbf{FEM solution}. The reader meshes the plate (taking
advantage of the two-fold symmetry to model only a quarter of the
plate), runs the analysis on three progressively refined meshes,
and records the peak stress at the hole on each mesh. The
mesh-convergence study should show convergence to within a few
percent of the analytical result on the finest mesh.

\textbf{Reconciliation}. The reader writes a one-page section
identifying:

\begin{itemize}[leftmargin=*]
  \item The peak stress from the analytical solution.
  \item The peak stress from the converged FEM solution.
  \item The percentage difference and its sources (finite-width
    correction, mesh resolution at the hole, element type and
    order).
  \item The mesh-convergence study (a plot of peak stress versus
    inverse element size at the hole) and the inferred converged
    value.
  \item The implication for engineering practice: at what mesh
    resolution would the FEM peak stress have been wrong by more
    than $10\,\%$?
\end{itemize}

\textbf{Deliverable}. The analytical derivation, the FEM model
files (mesh, input deck, output), the mesh-convergence plot, the
peak-stress comparison, and a short conclusion paragraph on the
reconciliation. The reader is encouraged to use an open-source
FEM package (CalculiX, Code\_Aster, FEniCS) so the workflow is
reproducible.

\textbf{Time}. Two to four hours for the analytical solution,
four to eight hours for the FEM setup and mesh-convergence study,
two to four hours of write-up. The project is suitable as a
weekend exercise once the reader is comfortable with the FEM
software.
\end{project}

\section{Exercises}

The exercises mix calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A simply supported square steel plate of side $1\,\si{\meter}$ and
thickness $10\,\si{\milli\meter}$ ($E = 200\,\text{GPa}$, $\nu =
0.3$) carries a uniform pressure of $5\,\text{kPa}$. Compute the
flexural rigidity $D$ and the centre deflection.

\paragraph{Solution sketch.} $D = Eh^3/[12(1-\nu^2)] = 200\times 10^9
\times (0.01)^3/[12\times 0.91] = 1.83\times 10^4\,\text{N}\cdot\text{m}$.
For a simply supported square plate $w_{\max} = 0.00406\, q_0 a^4/D =
0.00406\times 5000\times 1/(1.83\times 10^4) = 1.11\times 10^{-3}\,
\text{m} \approx 1.1\,\si{\milli\meter}$. The ratio $w/h = 0.11 < 1$
confirms linear theory applies. This is Worked example 1.
\end{exercise}

\begin{exercise}
A clamped circular aluminium plate of radius $0.5\,\si{\meter}$
and thickness $4\,\si{\milli\meter}$ ($E = 70\,\text{GPa}$, $\nu =
0.33$) carries a uniform pressure of $2\,\text{kPa}$. Compute the
centre deflection and the maximum bending moment at the clamped
edge.

\paragraph{Solution sketch.} $D = 70\times 10^9\times (0.004)^3/
[12(1-0.33^2)] = 70\times 10^9\times 6.4\times 10^{-8}/10.69 =
419\,\text{N}\cdot\text{m}$. For a clamped circular plate $w_{\max} =
q_0 a^4/(64D) = 2000\times 0.5^4/(64\times 419) = 2000\times 0.0625/
26816 = 4.66\times 10^{-3}\,\text{m} \approx 4.7\,\si{\milli\meter}$.
The edge moment is $M_{rr} = -q_0 a^2/8 = -2000\times 0.25/8 =
-62.5\,\text{N}\cdot\text{m}/\text{m}$; the surface stress there is
$6|M|/h^2 = 6\times 62.5/(0.004)^2 = 23.4\,\text{MPa}$.
\end{exercise}

\begin{exercise}
A thin spherical pressure vessel of radius $0.6\,\si{\meter}$,
wall thickness $4\,\si{\milli\meter}$, carries an internal gauge
pressure of $2\,\text{MPa}$. Compute the membrane stress in the
wall.

\paragraph{Solution sketch.} A sphere has equal meridional and
circumferential membrane stress $\sigma = pR/(2h) = 2\times 10^6
\times 0.6/(2\times 0.004) = 1.2\times 10^6/0.008 = 1.5\times 10^8\,
\text{Pa} = 150\,\text{MPa}$. Both principal stresses are equal, so
the von Mises stress equals $150\,\text{MPa}$ as well; against a
$250\,\text{MPa}$ yield the margin is $1.67$.
\end{exercise}

\begin{exercise}
A thin cylindrical steel pressure vessel of inner radius $0.5\,
\si{\meter}$ and wall thickness $6\,\si{\milli\meter}$ carries an
internal gauge pressure of $1.5\,\text{MPa}$. Compute the hoop
stress, the axial stress, and the ratio of the two.

\paragraph{Solution sketch.} Hoop $\sigma_\theta = pR/h = 1.5\times
10^6\times 0.5/0.006 = 1.25\times 10^8\,\text{Pa} = 125\,\text{MPa}$.
Axial $\sigma_z = pR/(2h) = 62.5\,\text{MPa}$. Ratio $\sigma_\theta/
\sigma_z = 2$, the universal thin-cylinder result; the hoop governs.
\end{exercise}

\begin{exercise}
A conical steel shell of half-angle $30^{\circ}$ and uniform wall
thickness $5\,\si{\milli\meter}$ carries an internal gauge pressure
of $0.5\,\text{MPa}$. Compute the circumferential and meridional
membrane stresses at a slant distance of $1\,\si{\meter}$ from
the apex.

\paragraph{Solution sketch.} With $\tan 30^\circ = 0.577$, $s = 1\,
\si{\meter}$, $h = 0.005\,\si{\meter}$: circumferential $\sigma_\theta
= p s\tan\alpha/h = 0.5\times 10^6\times 1\times 0.577/0.005 = 5.77
\times 10^7\,\text{Pa} = 57.7\,\text{MPa}$. Meridional $\sigma_s =
p s\tan\alpha/(2h) = 28.9\,\text{MPa}$, half the circumferential. The
stresses grow linearly with $s$, so the base is the critical section.
\end{exercise}

\begin{exercise}
A long aluminium beam ($E = 70\,\text{GPa}$, $\nu = 0.33$) of
square cross-section $40 \times 40\,\si{\milli\meter}$ is loaded
in plane strain with an in-plane uniaxial stress of $100\,\text{MPa}
$. Compute the effective two-dimensional Young's modulus and
Poisson's ratio.

\paragraph{Solution sketch.} Plane strain uses $E^* = E/(1-\nu^2) =
70/(1-0.33^2) = 70/0.891 = 78.6\,\text{GPa}$ and $\nu^* = \nu/(1-\nu)
= 0.33/0.67 = 0.493$. The effective modulus is stiffer and the
effective Poisson's ratio approaches $0.5$ because the through-thickness
constraint prevents the lateral contraction that a free bar would
have; the constraint makes the material behave as if nearly
incompressible in the plane.
\end{exercise}

\begin{exercise}
A simply supported rectangular steel plate of dimensions $1.5\,
\si{\meter} \times 1\,\si{\meter}$ and thickness $8\,\si{\milli
\meter}$ carries a uniform pressure of $3\,\text{kPa}$. Use the
Navier double-sine series, retaining the four smallest terms
($m, n = 1, 3$), to compute the centre deflection. Compare to the
single-term approximation.

\paragraph{Solution sketch.} $D = 200\times 10^9\times (0.008)^3/
[12\times 0.91] = 9.38\times 10^3\,\text{N}\cdot\text{m}$. At the
centre $\sin(m\pi/2)\sin(n\pi/2) = \pm 1$ for odd $m,n$. With $a =
1.5$, $b = 1$, the term denominators $mn[(m/a)^2+(n/b)^2]^2$ are:
$(1,1)$ gives $1\times[0.444+1]^2 = 2.085$ with sign $+$; $(1,3)$
gives $3\times[0.444+9]^2 = 267.6$ sign $-$; $(3,1)$ gives $3\times
[4+1]^2 = 75$ sign $-$; $(3,3)$ gives $9\times[4+9]^2 = 1521$ sign
$+$. Sum $= 1/2.085 - 1/267.6 - 1/75 + 1/1521 = 0.4796 - 0.00374 -
0.01333 + 0.00066 = 0.4632$. Then $w = (16 q_0/\pi^6 D)\times 0.4632 =
(16\times 3000/(961\times 9380))\times 0.4632 = 5.33\times 10^{-3}
\times 0.4632 = 2.47\times 10^{-3}\,\text{m}$. The single term alone
gives $5.33\times 10^{-3}\times 0.4796 = 2.56\,\si{\milli\meter}$, so
the four-term value $2.47\,\si{\milli\meter}$ is about three percent
lower; the series converges quickly. Cross-check with
\repopath{code/navier_plate.py}.
\end{exercise}

\begin{exercise}
A long thin-walled aluminium cylinder of radius $0.8\,\si{\meter}$
and wall thickness $3\,\si{\milli\meter}$ is loaded by external
hydrostatic pressure. Estimate the buckling pressure from the
classical shell-buckling formula, then apply a knock-down factor
of $0.3$ for imperfection sensitivity. Compare to the
gauge-pressure equivalent of a $20\,\si{\meter}$ depth of water.

\paragraph{Solution sketch.} For a long cylinder under external
pressure the classical critical pressure is $p_{\text{cr}} = E h^3/
[4(1-\nu^2)R^3]$ (the long-cylinder ring formula). With $E = 70\times
10^9$, $h = 0.003$, $R = 0.8$, $\nu = 0.33$: $p_{\text{cr}} = 70\times
10^9\times 2.7\times 10^{-8}/[4\times 0.891\times 0.512] = 1890/1.825
= 1.04\times 10^3\,\text{Pa} \approx 1.0\,\text{kPa}$. After the $0.3$
knock-down, $\approx 0.31\,\text{kPa}$. The pressure at $20\,\si{\meter}$
of water is $\rho g h = 1000\times 9.81\times 20 = 1.96\times 10^5\,
\text{Pa} = 196\,\text{kPa}$, more than two orders of magnitude above
the buckling capacity: this thin cylinder would collapse almost
immediately at depth. It needs ring stiffeners or a far thicker wall.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the plate flexural rigidity $D = E h^3/[12(1-\nu^2)]$ from
the Kirchhoff kinematic assumptions and the plane-stress Hooke's
law. State which integral is responsible for the $h^3$ dependence.

\paragraph{Solution sketch.} The Kirchhoff kinematics give
$\epsilon_{xx} = -z\,\partial^2 w/\partial x^2$ (and similarly for
$yy$, $xy$): strains linear in $z$. Plane-stress Hooke's law makes
$\sigma_{xx} = [E/(1-\nu^2)](\epsilon_{xx}+\nu\epsilon_{yy})$, also
linear in $z$. The bending moment is the first moment $M_{xx} =
\int_{-h/2}^{h/2} z\,\sigma_{xx}\,dz$. Substituting, the integrand
carries $z\times z = z^2$, and $\int_{-h/2}^{h/2} z^2\,dz = h^3/12$.
That integral is the source of the cube: one factor of $h$ comes from
the strain's linear $z$ dependence, one from the moment arm $z$, and
one from the range of integration. The result is $M_{xx} = -D(\partial^2
w/\partial x^2 + \nu\,\partial^2 w/\partial y^2)$ with $D = Eh^3/
[12(1-\nu^2)]$.
\end{exercise}

\begin{exercise}
Derive the plate bending equation $D \nabla^4 w = q$ from
transverse force equilibrium of an infinitesimal plate element
and the moment-curvature relations.

\paragraph{Solution sketch.} Transverse force balance on a $dx\,dy$
element gives $\partial Q_x/\partial x + \partial Q_y/\partial y + q =
0$. Moment balance about the two in-plane axes relates the shears to
the moment gradients: $Q_x = \partial M_{xx}/\partial x + \partial
M_{xy}/\partial y$ and $Q_y = \partial M_{yy}/\partial y + \partial
M_{xy}/\partial x$. Substituting eliminates the shears, leaving
$\partial^2 M_{xx}/\partial x^2 + 2\,\partial^2 M_{xy}/\partial x
\partial y + \partial^2 M_{yy}/\partial y^2 = -q$. Now insert the
moment-curvature relations $M_{xx} = -D(w_{,xx}+\nu w_{,yy})$,
$M_{yy} = -D(w_{,yy}+\nu w_{,xx})$, $M_{xy} = -D(1-\nu)w_{,xy}$. The
$\nu$ terms cancel in pairs and the twist term supplies the cross
derivative, collecting to $D(w_{,xxxx} + 2w_{,xxyy} + w_{,yyyy}) = q$,
i.e.\ $D\nabla^4 w = q$.
\end{exercise}

\begin{exercise}
Derive the membrane stresses in a thin spherical pressure vessel
under internal pressure from force equilibrium on a hemisphere.
State the assumption about wall thickness that the derivation
uses.

\paragraph{Solution sketch.} Cut the sphere along a diameter. The
internal pressure $p$ acting on the projected circular area $\pi R^2$
gives a net axial force $p\pi R^2$ pushing the two halves apart. The
wall resists it over the cut circumference $2\pi R$ with thickness $h$,
carrying membrane stress $\sigma$: resisting force $= \sigma\times 2\pi
R h$. Equilibrium $p\pi R^2 = \sigma\,2\pi R h$ gives $\sigma = pR/(2h)$.
By symmetry every great circle gives the same result, so the
meridional and circumferential stresses are equal. The assumption is
that $h \ll R$, so the stress is uniform through the wall (membrane,
no through-thickness bending) and $R$ is taken as the midsurface
radius.
\end{exercise}

\begin{exercise}
Derive the hoop and axial stresses in a thin cylindrical pressure
vessel from force equilibrium on a half-cylinder slice and on a
circular end cap, respectively.

\paragraph{Solution sketch.} \emph{Hoop:} slice a length $L$ of the
cylinder along a diametral plane. The pressure on the projected
rectangle $2RL$ gives force $p\,2RL$; the two longitudinal cut edges
(each $Lh$) resist with hoop stress $\sigma_\theta$: resisting force
$2\sigma_\theta L h$. Equilibrium $p\,2RL = 2\sigma_\theta Lh$ gives
$\sigma_\theta = pR/h$. \emph{Axial:} cap the cylinder and balance the
end thrust $p\pi R^2$ against the axial stress over the wall
cross-section $2\pi R h$: $p\pi R^2 = \sigma_z\,2\pi R h$, so $\sigma_z
= pR/(2h)$. Hence $\sigma_\theta = 2\sigma_z$.
\end{exercise}

\begin{exercise}
Show that, for an isotropic linear-elastic material, the
constitutive law $\sigma_{ij} = C_{ijkl} \epsilon_{kl}$ has only
two independent elastic constants, by enumerating the symmetries
of $C_{ijkl}$.

\paragraph{Solution sketch.} Start from $81$ components. Stress
symmetry $\sigma_{ij}=\sigma_{ji}$ forces $C_{ijkl}=C_{jikl}$ (minor
symmetry in $ij$), strain symmetry forces $C_{ijkl}=C_{ijlk}$ (minor
in $kl$); together these reduce $81$ to $36$. The existence of a
strain-energy potential gives the major symmetry $C_{ijkl}=C_{klij}$,
reducing $36$ to $21$ (the symmetric $6\times 6$ Voigt matrix).
Imposing full isotropy, the tensor must be invariant under all
rotations, and the only isotropic fourth-rank tensors with the right
symmetries are built from $\delta_{ij}\delta_{kl}$ and $(\delta_{ik}
\delta_{jl}+\delta_{il}\delta_{jk})$. Their two coefficients are the
Lam{\'e} parameters $\lambda$ and $\mu$, giving $C_{ijkl}=\lambda
\delta_{ij}\delta_{kl}+\mu(\delta_{ik}\delta_{jl}+\delta_{il}
\delta_{jk})$: two independent constants.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the central deflection of a $2 \,\text{m} \times 2\,\text{m}$
glass window pane, $6\,\si{\milli\meter}$ thick, simply supported
on all four edges, under a uniform wind pressure of $500\,\text{Pa}$.
State assumptions about glass elastic properties and check that
the deflection is within the linear-theory range.

\paragraph{Reference estimate.} Take glass $E \approx 70\,\text{GPa}$,
$\nu \approx 0.22$. $D = 70\times 10^9\times (0.006)^3/[12(1-0.048)] =
70\times 10^9\times 2.16\times 10^{-7}/11.42 = 1.32\times 10^3\,
\text{N}\cdot\text{m}$. For a simply supported square plate $w_{\max}
= 0.00406\,q_0 a^4/D = 0.00406\times 500\times 16/1320 = 0.0246\,
\text{m} \approx 25\,\si{\milli\meter}$. The ratio $w/h = 25/6 \approx
4 > 1$, so the pane is already past the linear-theory range and into
membrane action; the true deflection is smaller than $25\,\si{\milli
\meter}$ but the linear estimate flags that a $6\,\si{\milli\meter}$
pane spanning $2\,\si{\meter}$ is marginal and would normally be
thicker or smaller in practice.
\end{exercise}

\begin{exercise}
Estimate the burst pressure of a thin-walled aluminium soda can
($R \approx 33\,\si{\milli\meter}$, wall thickness $\approx 0.1\,
\si{\milli\meter}$, ultimate tensile strength $\approx 250\,\text{
MPa}$). Compare to a typical internal pressure ($\sim 300\,\text{
kPa}$) for a carbonated drink.

\paragraph{Reference estimate.} Treat the can wall as a thin cylinder;
the hoop stress governs, $\sigma_\theta = pR/h$. Setting $\sigma_\theta
= \sigma_{\text{ult}} = 250\,\text{MPa}$ and solving for the burst
pressure: $p_{\text{burst}} = \sigma_{\text{ult}}\,h/R = 250\times 10^6
\times 1\times 10^{-4}/0.033 = 2.5\times 10^4/0.033 = 7.6\times 10^5\,
\text{Pa} \approx 0.76\,\text{MPa}$. The carbonation pressure of about
$0.3\,\text{MPa}$ is roughly a factor of $2.5$ below burst, which
matches the lived experience that a shaken can can fail but a normal
one does not. The estimate also explains why the can ends are domed:
the flat-end stress would govern instead.
\end{exercise}

\begin{exercise}
Estimate the natural frequency of a clamped circular drumhead of
radius $0.15\,\si{\meter}$, thickness $0.2\,\si{\milli\meter}$,
material density $1400\,\si{\kilogram\per\meter\cubed}$,
Young's modulus $\sim 3\,\text{GPa}$.

\paragraph{Reference estimate.} The fundamental frequency of a clamped
circular plate is $f = (\lambda^2/2\pi)\sqrt{D/(\rho h a^4)}$ with
$\lambda^2 = 10.21$ for the lowest mode. $D = 3\times 10^9\times
(2\times 10^{-4})^3/[12(1-0.16)] \approx 3\times 10^9\times 8\times
10^{-12}/10.08 = 2.38\times 10^{-3}\,\text{N}\cdot\text{m}$ (using
$\nu \approx 0.4$). Mass per area $\rho h = 1400\times 2\times 10^{-4}
= 0.28\,\si{\kilogram\per\meter\squared}$. Then $D/(\rho h a^4) =
2.38\times 10^{-3}/(0.28\times 0.15^4) = 2.38\times 10^{-3}/(0.28\times
5.06\times 10^{-4}) = 16.8\,\text{s}^{-2}$, $\sqrt{\cdot} = 4.1$. So
$f = (10.21/6.28)\times 4.1 \approx 6.7\,\text{Hz}$ if read as plate
bending. A real drumhead is membrane-tensioned, which raises this
substantially; the bending estimate is a lower bound and shows the
plate stiffness alone gives only a few hertz, so the audible pitch
must come from tension, not flexural rigidity.
\end{exercise}

\begin{exercise}
Estimate the buckling pressure of a thin steel hemisphere of
radius $1\,\si{\meter}$ and wall thickness $5\,\si{\milli\meter}$
under external pressure. State the linear-theory critical
pressure and the realistic value after the imperfection knock-down.

\paragraph{Reference estimate.} $p_{\text{cr}} = 2E h^2/[R^2\sqrt{3
(1-\nu^2)}]$ with $E = 200\,\text{GPa}$, $h = 0.005$, $R = 1$, $\nu =
0.3$. Numerator $2\times 200\times 10^9\times 2.5\times 10^{-5} =
1.0\times 10^7$; denominator $\sqrt{3\times 0.91} = 1.65$, so
$p_{\text{cr}} = 6.05\,\text{MPa}$ (linear). After a $0.3$ knock-down,
$\approx 1.8\,\text{MPa}$. This is Worked example 5; the factor of
three between theory and capacity is the imperfection penalty.
\end{exercise}

\begin{exercise}
Estimate the peak stress concentration factor at the corner of a
square cutout in a flat plate under uniaxial tension. State why
the geometric singularity produces an unbounded prediction in
the elastic-theory limit and what design discipline replaces
the peak-stress check.

\paragraph{Reference estimate.} A truly sharp re-entrant corner (zero
radius) has a stress that grows as $r^{-\lambda}$ with $\lambda \approx
0.46$ for a $90^\circ$ notch, so the elastic stress is unbounded as
$r\to 0$ and the stress-concentration factor is formally infinite. The
FEM peak rises without bound as the mesh refines, so there is no
converged number to report. The design discipline that replaces the
peak-stress check is fracture or notch mechanics: characterise the
notch by its stress-intensity factor or by a finite notch radius from
the manufacturing process (a real corner always has a small radius),
and apply a fatigue or fracture criterion. The practical estimate for
a fillet of radius $\rho$ at a corner is $K_t \approx 1 + 2\sqrt{t/\rho}$
for a notch of depth $t$, finite only because $\rho > 0$.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Using an open-source finite-element package, mesh a square plate
with a central circular hole. Apply uniaxial tension to two
opposite edges. Run the analysis on three meshes (coarse, medium,
fine) and tabulate the peak stress at the hole on each mesh.
Estimate the converged value by Richardson extrapolation.

\paragraph{Solution sketch.} The peak stress should rise from below
toward $K_t\sigma_\infty$ as the mesh refines; for a small hole and
remote tension $\sigma_\infty$, the converged value is $3\sigma_\infty$.
With three results such as the $128$, $139$, $146\,\text{MPa}$ sequence
of \repopath{data/mesh_convergence.csv} at refinement ratio $r = 2$,
Richardson extrapolation gives $p = \ln[(f_1-f_2)/(f_2-f_3)]/\ln r =
\ln(11/7)/\ln 2 \approx 0.65$ and $f_{\text{exact}} = f_3 + (f_3-f_2)/
(r^p-1) \approx 146 + 7/0.57 \approx 158\,\text{MPa}$ from this coarse
triple; refining further tightens it toward the analytical $150\,
\text{MPa}$. The key reported quantities are the converged value, the
observed order, and the grid-convergence index; see
\repopath{code/richardson_convergence.py}.
\end{exercise}

\begin{exercise}
Implement a hand-coded two-dimensional plane-stress finite-element
solver using linear triangular elements. Verify on a uniaxial
tension test against the analytical solution.

\paragraph{Solution sketch.} The solver needs four pieces: the
plane-stress material matrix $\mathbf{D} = [E/(1-\nu^2)]\,[[1,\nu,0],
[\nu,1,0],[0,0,(1-\nu)/2]]$; the constant strain-displacement matrix
$\mathbf{B}$ built from the triangle's nodal coordinates; the element
stiffness $K_e = t\,A\,\mathbf{B}^{\mathsf T}\mathbf{D}\,\mathbf{B}$;
and assembly plus a constrained solve $K\vec{u}=\vec{F}$. The patch
test loads a single (or two-triangle) square in uniaxial tension and
checks that the recovered strain equals $\sigma/E$ and the stress
field is uniform. A complete reference implementation that passes this
test is \repopath{code/cst_fem.py}; a correct solver reproduces $\sigma
= E\epsilon$ to machine precision on the patch test, which is the
minimum bar an element formulation must clear.
\end{exercise}

\begin{exercise}
Using a finite-element package, model a cantilevered thin square
plate of side $200\,\si{\milli\meter}$ and thickness $2\,\si{
\milli\meter}$ as both (a) a shell element model and (b) a
single-layer brick element model. Compare the tip deflection
under a $10\,\text{N}$ tip load with the analytical Kirchhoff
plate prediction. Identify the source of any discrepancy in the
brick-element model.

\paragraph{Solution sketch.} The Kirchhoff cantilever-strip estimate
treats the plate as a beam of width $b = 0.2\,\si{\meter}$, second
moment $I = bh^3/12 = 0.2\times (0.002)^3/12 = 1.33\times 10^{-10}\,
\text{m}^4$, length $L = 0.2$, $E = 200\,\text{GPa}$: tip deflection
$\delta = PL^3/(3EI) = 10\times 0.008/(3\times 200\times 10^9\times
1.33\times 10^{-10}) = 0.08/79.8 = 1.0\times 10^{-3}\,\text{m}\approx
1\,\si{\milli\meter}$. The shell-element model should match this to a
few percent. The single-layer brick model returns roughly half to a
third of it because of shear locking
(figure~\ref{fig:vol03ch09:shear-locking}): one linear element through
the thickness cannot bend and manufactures parasitic shear strain. The
cure is shell elements, several quadratic bricks through the thickness,
or reduced integration.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a flat steel plate floor of a small workshop, $3\,\si{
\meter} \times 4\,\si{\meter}$, simply supported on its four
edges by perimeter beams, to carry a uniform live load of $5\,
\text{kPa}$. Specify the plate thickness and verify the
small-deflection assumption.

\paragraph{Model-answer sketch.} The short span $a = 3\,\si{\meter}$
governs. Set a serviceability deflection limit, say $w \le L/250 =
12\,\si{\milli\meter}$. For a rectangular plate near the one-way limit
the deflection coefficient is around $0.01$, so $w \approx 0.01\,q_0
a^4/D$ and the required $D \ge 0.01\times 5000\times 81/0.012 = 3.4\times
10^6\,\text{N}\cdot\text{m}$. With steel $D = Eh^3/[12(1-\nu^2)] =
1.83\times 10^{10}h^3$, this needs $h^3 \ge 1.85\times 10^{-4}$, so
$h \ge 57\,\si{\milli\meter}$, which is implausibly thick for a bare
plate. The honest answer is that a flat plate alone is the wrong
structure for a $3\times 4\,\si{\meter}$ floor; the design adds
stiffening ribs or a grillage so each plate panel spans under a metre,
or uses a composite deck. The exercise teaches that the thickness-cubed
penalty makes large unstiffened plates uneconomic, and verifies $w/h$
is small only after stiffening reduces the panel span.
\end{exercise}

\begin{exercise}
Design a thin-walled cylindrical steel pressure vessel of inner
radius $0.4\,\si{\meter}$ to carry an internal pressure of $4\,
\text{MPa}$ with a safety factor of $3$ on hoop stress. Specify
the wall thickness and a code-style longitudinal weld detail.

\paragraph{Model-answer sketch.} Allowable stress $\sigma_{\text{allow}}
= \sigma_Y/3$. Taking $\sigma_Y = 250\,\text{MPa}$, $\sigma_{\text{
allow}} = 83.3\,\text{MPa}$. Hoop governs: $h = pR/\sigma_{\text{allow}}
= 4\times 10^6\times 0.4/83.3\times 10^6 = 19.2\,\si{\milli\meter}$;
specify $h = 20\,\si{\milli\meter}$. The longitudinal weld sees the
full hoop stress, so it is the governing seam; a code design specifies
a full-penetration double-welded butt joint with a joint efficiency
$E_j$ (ASME applies $E_j \approx 0.85$ for spot-radiographed,
$1.0$ for fully radiographed), increasing the required thickness to
$h = pR/(\sigma_{\text{allow}}E_j)$. With $E_j = 0.85$, $h = 22.6\,
\si{\milli\meter}$; specify $24\,\si{\milli\meter}$ and full
radiography of the longitudinal seam. The circumferential seam, at
half the stress, can use a lower joint efficiency.
\end{exercise}

\begin{exercise}
Design a hemispherical pressure-vessel end cap to match a
cylindrical body of inner radius $0.4\,\si{\meter}$, internal
pressure $4\,\text{MPa}$, with the same safety factor. Specify
the end-cap thickness and note the comparison to the cylindrical
wall thickness.

\paragraph{Model-answer sketch.} The hemispherical cap carries $\sigma
= pR/(2h)$, so $h_{\text{cap}} = pR/(2\sigma_{\text{allow}}) = 4\times
10^6\times 0.4/(2\times 83.3\times 10^6) = 9.6\,\si{\milli\meter}$;
specify $10\,\si{\milli\meter}$. This is half the cylinder wall
($\approx 20\,\si{\milli\meter}$ from the previous exercise), the
quantitative statement that a hemispherical head is the most efficient
closure. Note the discontinuity: the thin cap meets the thicker
cylinder at the junction, where the mismatch in radial growth under
pressure induces a bending boundary layer. A real design either tapers
the thickness transition over a few wall thicknesses or uses a knuckle
radius (a torispherical or ellipsoidal head) to soften the
discontinuity.
\end{exercise}

\begin{exercise}
Design a finite-element mesh strategy for analysing the stress
concentration at a fillet between a shaft and its mounting
flange, with the fillet radius equal to one-tenth the shaft
radius. Specify the element type, the minimum element size at
the fillet, and the criteria for accepting the converged peak
stress.

\paragraph{Model-answer sketch.} Use quadratic (second-order)
elements: tetrahedra if the geometry forces an unstructured mesh,
hexahedra if a swept mesh is possible, because linear elements
under-predict the fillet peak. The element size at the fillet should
resolve the radius with several elements across it; with a fillet
radius $\rho = R_{\text{shaft}}/10$, set the minimum element size to
about $\rho/4$ to $\rho/6$, giving four to six elements through the
fillet arc. Accept the peak stress only after a mesh-convergence study
of three meshes shows the peak changing by less than a chosen tolerance
(say $2\%$) between the two finest meshes, with the convergence rate
consistent with the element order. Because a fillet has a finite
radius, the peak is bounded and converges (unlike a sharp corner), so
a converged value exists and is the design stress; apply the fatigue
check from chapter 8 to it.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A finite-element analysis of a bolted bracket predicts a peak
stress of $400\,\text{MPa}$ at a node on a sharp re-entrant
corner. The mesh refinement is doubled; the predicted peak rises
to $580\,\text{MPa}$. Identify the underlying issue and the
diagnostic that catches it. State the engineering criterion that
should replace the peak-stress check.

\paragraph{Solution sketch.} The peak stress rising with mesh
refinement is the signature of a geometric stress singularity at the
sharp re-entrant corner: the elastic stress is unbounded there, so the
FEM peak diverges and never converges. The diagnostic is the
mesh-convergence study itself, which shows the peak increasing rather
than stabilising. The corner stress is not a physical failure
criterion; the criterion that replaces it is a fracture- or
notch-mechanics check (stress-intensity factor, the structural-stress
or notch-stress method, or a finite fillet radius from the actual
manufactured part). A real corner has a small radius; modelling that
radius bounds the stress and restores convergence.
\end{exercise}

\begin{exercise}
A thin spherical pressure vessel is observed to buckle inward
under external vacuum loading at a pressure differential
significantly below the linear-theory critical pressure. Identify
the dominant cause and state the design discipline that should
have anticipated the failure.

\paragraph{Solution sketch.} The dominant cause is imperfection
sensitivity: a real spherical shell buckles at $20$ to $40\%$ of the
linear-theory critical pressure because small geometric imperfections
(a dimple, a flat spot, out-of-roundness) destabilise the membrane
equilibrium. The design discipline that anticipates it is to apply a
knock-down factor calibrated to test data, never to design against the
linear value; design codes (ASME Section VIII, the NASA shell-buckling
monographs) supply the reduced capacities. The fix is a thicker wall,
ring or stringer stiffeners, or a tighter manufacturing tolerance on
out-of-roundness.
\end{exercise}

\begin{exercise}
A cantilevered FEM model of a thin metal panel returns a tip
deflection that is approximately half of the closed-form
prediction from Kirchhoff plate theory. The model uses linear
brick elements with one element through the thickness. Identify
the numerical artifact and the corrective action.

\paragraph{Solution sketch.} The artifact is shear locking. A single
linear brick through the thickness has straight edges and cannot
represent the curvature of pure bending; it develops a parasitic
transverse shear strain that absorbs energy and makes the panel report
far too stiff, hence a deflection roughly half the Kirchhoff value
(figure~\ref{fig:vol03ch09:shear-locking}). The corrective action is to
switch to shell elements, use several quadratic bricks through the
thickness, or use reduced-integration solid elements with hourglass
control. A through-thickness mesh-refinement study would also expose it.
\end{exercise}

\begin{exercise}
A wooden floor beam analysed as isotropic with the wood's
longitudinal Young's modulus predicts deflections half those
actually observed. Identify the modelling error and the
corrective material model.

\paragraph{Solution sketch.} The error is treating wood as isotropic
with its largest (longitudinal, along-grain) modulus, which overstates
the bending stiffness when the beam is loaded so that the radial or
tangential modulus matters, or when shear deformation along the grain
is significant. Wood is orthotropic: the longitudinal modulus exceeds
the radial and tangential moduli by a factor of $20$ or more, and the
shear moduli are smaller still. The corrective model is an orthotropic
constitutive law with the grain direction aligned to the beam axis and
the radial and tangential moduli and the relevant shear modulus
entered correctly; design then follows the grain-direction allowables
(NDS or Eurocode 5). Doubling of the predicted-versus-observed
deflection is consistent with a missing shear-deflection contribution
and an overstated effective modulus.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Re-read the NASA Optical Systems Failure Report on the Hubble
primary mirror \cite{acc:nasa-hubble-optical-systems-1990}.
Identify the parallel between the metrology-chain failure in
Hubble fabrication and the VVUQ failure modes of section 9.7.
State the specific verification step that, if applied to the
reflective null corrector, would have caught the error before
the mirror was polished to its final figure.

\paragraph{Model-answer sketch.} The parallel is the unverified test
instrument. In Hubble the reflective null corrector was treated as the
authoritative figure measurement, exactly as a beginner treats an FEM
result as authoritative; both are predictions whose trustworthiness
depends on a verification chain that was skipped. The specific step
that would have caught it is independent verification of the test
instrument: an end-to-end check of the null corrector against a second,
independent metrology method (the auxiliary refractive null corrector
and the inverse-null test that did show discrepancies but were set
aside). In FEM terms this is code verification (does the tool solve the
intended problem) plus anomaly resolution (take every discrepancy
between the authoritative result and an independent check seriously).
The discipline is to verify the instrument, not only the part it
measures.
\end{exercise}

\begin{exercise}
A thin-walled aluminium aircraft fuselage panel, designed by
finite-element analysis as a flat plate in plane stress, develops
visible buckles under combined cabin pressurisation and bending
during pre-service ground testing. Argue, in $200$ to $300$ words,
which class of FEM error (modelling, discretisation, interpretation)
is most likely responsible and what verification should have
caught it.

\paragraph{Model-answer sketch.} The most likely class is a modelling
error: a linear static plane-stress analysis cannot predict
out-of-plane buckling, because buckling is a stability (eigenvalue)
problem, not a stress problem. The flat-plate plane-stress idealisation
discards the very degree of freedom (transverse displacement) in which
the panel buckles. A panel can be well below its yield stress and still
buckle. The verification that should have caught it is a linear
buckling (eigenvalue) analysis followed by a geometrically nonlinear
post-buckling analysis with an imperfection seeded, and validation
against a coupon or sub-component compression test. A converged stress
plot from a linear static run says nothing about stability; the
discipline is to recognise the loading as stability-driven and to run
the buckling analysis the static model omitted. A good answer names the
class as modelling, identifies the missing physics as stability, and
prescribes a buckling eigenvalue analysis plus imperfection-sensitive
nonlinear check.
\end{exercise}

\begin{exercise}
A thin spherical storage tank for liquefied gas, designed against
the linear-theory buckling pressure with a factor of safety of
$1.5$, collapses inward when a vapour-return valve fails closed
during pump-out. Argue, in $200$ to $300$ words, which factor of
the design was responsible for the failure and what design
discipline catches the imperfection-sensitivity issue.

\paragraph{Model-answer sketch.} The responsible factor is designing
against the linear-theory buckling pressure with only a $1.5$ safety
factor. Real shells buckle at $20$ to $40\%$ of the linear value, so a
$1.5$ factor against the linear capacity leaves an actual margin below
unity once imperfection sensitivity is accounted for; an external
pressure the design treated as within margin was in fact above the true
buckling capacity. The valve failing closed during pump-out created an
unanticipated external pressure differential (partial vacuum inside),
the loading the inward-buckling mode responds to. The discipline that
catches it is a calibrated knock-down factor on the linear buckling
load, per ASME Section VIII or the NASA shell-buckling monographs, plus
a vacuum-breaker or relief device sized so the credible valve-failure
pressure cannot exceed the knocked-down capacity. A good answer names
the linear buckling basis as the fault and the empirical knock-down
(with a vacuum breaker) as the cure.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer presents a FEM analysis of a structural
component to the design review meeting, with no mesh-convergence
study and no validation against any analytical or experimental
data. The peak stress is below the material's yield strength,
and the engineer recommends the design be approved. Write a
one-paragraph reply that identifies what is missing and what
verification the design review should require before approval.

\paragraph{Model-answer sketch.} What is missing is the entire
verification-and-validation chain: a single FEM run on a single mesh
gives a number whose error is unknown, and a peak stress below yield is
not a result if it has not been shown to be mesh-independent. The
review should require, before approval, a mesh-convergence study on the
quantity of interest with the observed convergence order reported; a
check that the peak is not at a geometric singularity (where no
converged value exists); validation against an analytical result or a
test for at least one representative case; and a proper code-based
design check applying the right factor of safety, not a bare
stress-versus-yield comparison. The reply should say plainly that
"below yield on one mesh" is a colourful screenshot, not an engineering
claim, and name the convergence study as the minimum gate.
\end{exercise}

\begin{exercise}
A senior colleague argues that commercial FEM software is
sufficiently mature that engineers no longer need to understand
the underlying mathematics. Write a one-paragraph reply that
identifies the specific knowledge a working FEM analyst must
retain and the failure modes that opaque software use admits.

\paragraph{Model-answer sketch.} The reply should grant that solvers
and element libraries are mature, then state what no maturity can
substitute for: the analyst must understand the governing physics
(stress versus stability, linear versus nonlinear, static versus
dynamic), the element-formulation pitfalls (shear and volumetric
locking, singularities at sharp corners, stress accuracy at Gauss
points versus nodes), the boundary-condition and load idealisations,
and the verification-and-validation discipline. The failure modes that
opaque use admits are exactly the ones in section 9.7: an isotropic
model of an anisotropic material, a linear static run where buckling
governs, a coarse mesh that under-predicts a concentration, a
singularity read as a failure stress, unit confusion, and a peak read
at an inaccurate node. Mature software runs all of these to convincing
colour plots. The knowledge the analyst retains is the knowledge of
when the software is solving the wrong problem.
\end{exercise}

\begin{exercise}
Identify a thin-walled or plate-like structural component in
your immediate environment whose loading you have never explicitly
analysed (a metal cabinet panel, a glass tabletop, a sheet-metal
appliance casing, a thin polystyrene shelf). State, in $200$ to
$300$ words, the loads it carries, the dominant deformation mode,
the relevant simplified theory (plate, shell, membrane, or
three-dimensional elasticity), and the failure mode you would
inspect for first.

\paragraph{Model-answer sketch.} A good answer picks one component and
carries it through the chapter's framework. For a sheet-metal appliance
casing panel, for instance: the loads are its own weight, the press
fit of fasteners, and incidental hand pressure; the dominant
deformation mode is plate bending of the unsupported spans between
stiffening ribs or fold lines; the relevant theory is Kirchhoff plate
theory for the flat panels and shell or fold geometry where the metal
is bent (a fold is a stiffener because it gives the sheet local
curvature). The failure mode to inspect first is local buckling or
oil-canning of the largest unsupported flat span under modest pressure,
since a thin flat panel is far more flexible than the same metal folded.
The answer is judged on correct identification of the deformation mode
and the matching simplified theory, not on the specific object.
\end{exercise}
