\chapter{Energy methods}
\label{vol03:ch05}

\epigraph{No figures will be found in this work. The methods I expound
require neither constructions, nor geometrical or mechanical
arguments, but solely algebraic operations, subject to a regular and
uniform procedure.}{Joseph-Louis Lagrange, preface to \emph{M\'ecanique
Analytique} (1788), in the Boissonnade and Vagliente translation
\cite[p.~7]{hist:lagrange1788}.}

\chapmeta{Half-life: durable. Archetypes: balance, optimisation. Exercise target: 30. Project track: derivation plus simulation (hybrid). Hazard class: none. Reader path default: standard, with sections explicitly tagged. Named cases: Tacoma Narrows Bridge (1940).}

\noindent
Chapter 4 of this volume treated dynamics in the Newtonian form: a
free-body diagram for every body, a vector equation $\vec{F} = m\vec{a}$
for the translation of every centre of mass, and a moment equation for
every rotation. The method is correct and general. It is also,
for systems with many bodies, many constraints, or constraints of
complicated geometry, computationally heavy: writing the equations is
mechanical but error-prone, and the unknown internal forces between
connected bodies enter equations that the engineer then has to
eliminate. The energy-method reformulation of mechanics replaces vector
balances with scalar quantities and turns the equations of motion into
consequences of a single variational principle. The reformulation is
the work of Lagrange (1788) and Hamilton (1834); its mature engineering
form is the working tool of theoretical mechanics, structural
dynamics, robotics, and every field in which mechanical systems with
many degrees of freedom need to be analysed without bookkeeping
error. The chapter builds the scalar apparatus, works it on the
pendulum, the double pendulum, and the bead on a rotating hoop, and
closes on the failure mode the method makes possible: an analysis whose
generalised coordinates are too few to represent the motion that
actually destroys the system, the mode that brought down the Tacoma
Narrows bridge.

\section{Energy as a unifying currency}\pathtag{core}

Energy in mechanics is the integral of force through displacement,
established in chapter 4 as the work done by a force on a particle.
For a conservative force the integral depends only on the endpoints,
and the work is the change in a scalar potential energy associated
with the force. Kinetic energy is a scalar quadratic form in the
velocity. Conservation of mechanical energy in a system with only
conservative forces is the statement that the sum of kinetic and
potential energy is constant in time.

The reformulation moves three quantities to centre stage.

\engterm{Kinetic energy} $T$. For a system of particles,
\[
T = \sum_i \tfrac{1}{2} m_i \vec{v}_i \cdot \vec{v}_i,
\]
a scalar function of the velocities. For a rigid body in plane motion,
$T = \tfrac{1}{2} m v_G^2 + \tfrac{1}{2} I_G \omega^2$, as derived in
chapter 4. In all cases $T$ is a quadratic, positive-semi-definite
function of the velocities, with $T = 0$ if and only if every particle
is at rest.

\engterm{Potential energy} $U$. For conservative forces, $U$ is a
scalar function of position such that $\vec{F}_i = -\nabla_i U$, with
$\nabla_i$ the gradient with respect to particle $i$'s position. The
gravitational potential near Earth's surface is $U = mgz$ per particle;
the spring potential is $U = \tfrac{1}{2}k x^2$ per spring;
electrostatic, magnetic-dipole, and many central-force potentials have
their own standard forms.

\engterm{Generalised force}. For non-conservative forces (friction,
external applied forces, time-dependent driving forces), the system
gains a quantity that has the dimensions of force times displacement
per unit virtual displacement of a generalised coordinate; it is the
generalised force, and we will define it precisely in section 5.3
below. The non-conservative forces are then read into the formalism
without disturbing the conservative part.

The unification is not yet visible in this list. Each of the three
quantities is a scalar; combinations of them generate vector force
equations after differentiation. The variational principle of section
5.5 will state that, among all conceivable trajectories from a given
initial state to a given final state at given times, the actual
trajectory is the one that makes the time-integral of $L = T - U$
stationary. The variational form is the unifying frame; the
Newton-equivalent equations of motion emerge from it without ever
writing a vector force balance.

The engineering payoff of the reformulation is twofold. First, the
working coordinates are not Cartesian particle positions but
\emph{generalised coordinates}: any set of independent quantities
that locate the system. A pendulum has one generalised coordinate
(the swing angle), not two Cartesian particle positions; a double
pendulum has two; a robot arm has as many as its joints. Constraints
that would appear as forces in the Newtonian formulation are absorbed
into the choice of coordinates and produce no force terms in the
Lagrangian. Second, the analysis is local in the sense that bookkeeping
becomes mechanical: write $T$, write $U$, take derivatives in a
standard pattern, and the equations of motion emerge. The
error-proneness of free-body bookkeeping is replaced by the
arithmetic of partial derivatives, which is more reliable.

\begin{principle}[Mechanics in scalar form]
A mechanical system's motion can be derived from two scalar
functions of the system's state, the kinetic energy $T$ and the
potential energy $U$, together with a specification of any
non-conservative generalised forces. The equations of motion are
consequences of a single variational principle on the action
integral. No vector force balance need be drawn, and no internal
constraint force need be solved for, provided the generalised
coordinates are chosen to incorporate the constraints.
\end{principle}

A picture sharpens the claim that scalars carry the dynamics. Take the
simplest conservative system, a small-amplitude pendulum, and watch its
two energies over one period. At the turning points the bob is
momentarily at rest and all its energy sits in the gravitational
potential; at the bottom of the swing the height is least and the speed
greatest, so the energy has moved entirely into the kinetic term. The
two quantities trade back and forth, twice per period, and their sum
holds flat because gravity is conservative and the pivot does no work.
Figure~\ref{fig:vol03ch05:energy-exchange} draws the exchange. The
energy method is the discipline of reading the dynamics off this kind
of scalar accounting rather than off a force diagram.

\input{volumes/03-force/ch05-energy-methods/figures/fig-energy-exchange}

A second picture fixes the relation between a potential and the force
it encodes. For a single coordinate the conservative force is the
negative slope of the potential, $F = -dU/dx$, so a potential drawn as
a curve carries the whole force law in its shape: equilibria are its
stationary points, the restoring force at any displacement is its local
slope, and the stiffness is its curvature. The spring potential
$U = \tfrac12 k x^2$ is the canonical well, drawn in
figure~\ref{fig:vol03ch05:spring-potential}; a stiffer spring is a
steeper well, and the bottom of the well is the stable equilibrium to
which the mass returns. Most of the analysis in this chapter is the
multi-coordinate generalisation of reading a curve: locate the
stationary points of an energy, examine the curvature there, and let
the equilibria and their stability follow.

\input{volumes/03-force/ch05-energy-methods/figures/fig-spring-potential}

\begin{estimation}
\textbf{Question.} A child on a $2.5\,\si{\meter}$ playground swing
starts from rest with the ropes horizontal (a $90^{\circ}$ amplitude)
and is brought to rest by a parent at the bottom of the first descent.
Estimate the impulse the parent must supply and the average force over a
half-second catch, and check whether energy methods or a force balance
is the shorter route.

\textbf{Working.} Take the child plus seat as a point mass of
$m \approx 25\,\si{\kilogram}$. The drop from horizontal ropes to the
bottom is the rope length, $h = L = 2.5\,\si{\meter}$. Energy
conservation gives the speed at the bottom without any path integration:
$\tfrac12 m v^2 = m g h$, so
\[
v = \sqrt{2 g h} = \sqrt{2 \cdot 9.81 \cdot 2.5}
\approx 7.0\,\si{\meter\per\second}.
\]
The momentum to be removed is $p = m v \approx 25 \cdot 7.0
= 175\,\si{\kilogram\meter\per\second}$, which is the impulse required.
Over a catch lasting $\Delta t \approx 0.5\,\si{\second}$ the average
force is
\[
F_\text{avg} \approx \frac{p}{\Delta t}
= \frac{175}{0.5} = 350\,\si{\newton},
\]
roughly the child's weight again ($mg \approx 245\,\si{\newton}$)
plus enough to decelerate, a few hundred newtons, demanding but
manageable for an adult. A force-balance route would have required
integrating the tension and the tangential weight component along the
quarter-circle arc to reach the same speed; energy conservation read it
off the height drop in one line. The number to carry away is the order
of magnitude: catching a swing at full amplitude is a several-hundred-
newton job, and the energy method is the faster path to it.
\end{estimation}

\section{Conservation of energy in mechanics}\pathtag{standard}

For a system whose forces are all conservative and whose constraints
do no work (a constraint is workless if its reaction force is
perpendicular to the permitted virtual displacement, which is true of
smooth pin joints, smooth surfaces, and inextensible ropes), the
total mechanical energy is constant in time:
\[
E = T + U = \text{constant}.
\]
The result follows from Newton's second law $\vec{F} = m\vec{a}$ and
the identity $\vec{a}\cdot d\vec{r} = d(\tfrac{1}{2} v^2)$, as derived
in chapter 4. For a system that includes non-conservative forces
(friction, drag, applied driving), energy is not conserved; the rate
of change of $T + U$ equals the rate at which non-conservative forces
do work on the system.

\begin{archetype}[Balance]
Conservation of energy is the mechanical instance of the balance
archetype: a quantity is neither created nor destroyed, so its
accumulation in a system equals the net flow across the boundary. The
same accounting governs conservation of mass in a control volume
(Volume IV), conservation of charge at a circuit node (Kirchhoff's
current law), conservation of momentum in a collision, and the first
law of thermodynamics, where the change in internal energy balances heat
added and work done. The recurring discipline is to draw a boundary,
list every inflow and outflow of the conserved quantity, and set the
accumulation equal to the net flux. In a conservative mechanical system
the boundary flux is zero and the balance collapses to $T + U =
\text{constant}$; the energy method generalises this from a constant to
a rate equation when non-conservative work crosses the boundary.
\end{archetype}

The principle of \engterm{virtual work} is the static-equilibrium
limit of energy methods. For a system in static equilibrium, any
virtual displacement consistent with the constraints (a hypothetical
infinitesimal motion that respects the geometric constraints but is
not necessarily a real motion of the system) produces zero virtual
work by the applied forces:
\[
\delta W = \sum_i \vec{F}_i \cdot \delta \vec{r}_i = 0.
\]
The sum is over the applied forces, not over the constraint forces,
because workless constraints produce no virtual work by definition.
The principle is equivalent to the statement that, in static
equilibrium with workless constraints, the applied forces' resultant
in every allowed direction of motion is zero.

For conservative applied forces, virtual work has the simpler form
\[
\delta U = 0,
\]
the variation of the potential energy under any allowed virtual
displacement vanishes. The applied forces' equilibrium is the
stationarity of $U$ under variations consistent with the constraints.
The result was established by Lagrange in the 1788 \emph{M\'ecanique
Analytique} as the generalisation of the equilibrium conditions to
arbitrary mechanical systems \cite{hist:lagrange1788}.

\textbf{Worked example: the lever by virtual work.} A rigid lever
pivots without friction at a fulcrum, with an applied force $F_1$ at
distance $a$ on one side and a resisting force $F_2$ at distance $b$ on
the other (figure~\ref{fig:vol03ch05:virtual-work}). The Newtonian
route writes a moment balance about the fulcrum and, if the full state
is wanted, a force balance that introduces the unknown pin reaction.
The virtual-work route skips the reaction. Give the lever a virtual
rotation $\delta\phi$ about the fulcrum, the one motion the pin
constraint permits. The two ends move through $\delta r_1 = a\,\delta\phi$
and $\delta r_2 = b\,\delta\phi$. The pin reaction does no virtual work
because the fulcrum point does not move. Setting the total virtual work
of the applied forces to zero,
\[
\delta W = F_1\,a\,\delta\phi - F_2\,b\,\delta\phi = 0
\quad\Longrightarrow\quad
F_1 a = F_2 b,
\]
the law of the lever, with the reaction never computed. The economy is
the same one the Lagrangian formulation delivers in dynamics: workless
constraints drop out, and a scalar balance survives.

\input{volumes/03-force/ch05-energy-methods/figures/fig-virtual-work}

\textbf{Worked example: speed from energy conservation.} A block of
mass $m$ released from rest slides without friction down a smooth track
and drops a height $h$. No force other than gravity does work (the
normal force is perpendicular to the motion and so is workless), so
$T + U$ is constant. Taking the release point as the height datum,
$0 + mgh = \tfrac12 m v^2 + 0$ at the bottom, hence
\[
v = \sqrt{2 g h},
\]
independent of the mass and of the shape of the track. The same result
extracted by Newton's method would require resolving the weight along a
track whose slope changes continuously and integrating the tangential
acceleration over the path; energy conservation reads the answer off
the endpoints. This is the recurring trade the chapter develops: a
scalar bookkeeping of energy replaces a vector integration along a
path, and the path's geometry, which dominates the Newtonian
calculation, drops out when only the endpoints matter.

D'Alembert's principle extends virtual work to dynamics: for a system
not in static equilibrium, the applied forces minus the inertial
``forces'' $-m\vec{a}$ produce zero virtual work for any virtual
displacement consistent with the constraints:
\[
\sum_i (\vec{F}_i - m_i \vec{a}_i) \cdot \delta \vec{r}_i = 0.
\]
The inertial term $-m\vec{a}$ is sometimes called a pseudo-force or
an inertial force; its appearance in the equation captures Newton's
second law's content as a special case of virtual work. The
formulation is the bridge from Newtonian dynamics to Lagrangian
mechanics: substituting $T$ and $U$ for the kinematic and force terms
turns d'Alembert's principle into the Euler-Lagrange equations of
the next section.

\subsection{The work-energy theorem and path-independence}

The scalar quantities of section 5.1 rest on a single identity carried
over from chapter 4, and the identity is worth deriving in full because
every later result inherits its structure. Take one particle of mass $m$
moving under a net force $\vec{F}$. Newton's second law is $\vec{F} =
m\,d\vec{v}/dt$. Form the dot product of both sides with the velocity
$\vec{v} = d\vec{r}/dt$,
\[
\vec{F}\cdot\vec{v} = m\,\frac{d\vec{v}}{dt}\cdot\vec{v}.
\]
The right side is a perfect time derivative, since
\[
\frac{d}{dt}\Bigl(\tfrac12\,\vec{v}\cdot\vec{v}\Bigr)
= \tfrac12\Bigl(\frac{d\vec{v}}{dt}\cdot\vec{v}
+ \vec{v}\cdot\frac{d\vec{v}}{dt}\Bigr)
= \frac{d\vec{v}}{dt}\cdot\vec{v},
\]
so $\vec{F}\cdot\vec{v} = \dfrac{d}{dt}\bigl(\tfrac12 m v^2\bigr) =
\dfrac{dT}{dt}$. The left side is the rate at which the force does work.
Multiplying through by $dt$ and integrating from an initial state $1$ to
a final state $2$,
\[
\int_1^2 \vec{F}\cdot d\vec{r} = T_2 - T_1,
\]
the \engterm{work-energy theorem}: the work done by the net force on a
particle equals the change in its kinetic energy. The result holds for
any force whatever, conservative or not, because nothing in the
derivation assumed a potential. It is Newton's second law integrated
once along the path, with the vector equation collapsed to a scalar
balance.

The theorem becomes the conservation law of section 5.2 only when the
force is conservative. A force is \engterm{conservative} when its work
between two points is the same for every path joining them, which is
equivalent to the line integral around every closed loop vanishing,
\[
\oint \vec{F}\cdot d\vec{r} = 0.
\]
Figure~\ref{fig:vol03ch05:path-independence} draws the property: two
routes from one point to another carry the same work. The closed-loop
condition is equivalent, by Stokes' theorem, to $\nabla\times\vec{F} =
\vec{0}$ everywhere in a simply connected region, and a curl-free field
is the gradient of a scalar. We name that scalar the potential energy
$U$ and fix its sign so that $\vec{F} = -\nabla U$. The work of a
conservative force is then a difference of potential values,
\[
\int_1^2 \vec{F}\cdot d\vec{r}
= -\int_1^2 \nabla U\cdot d\vec{r}
= -(U_2 - U_1) = U_1 - U_2,
\]
the integral of an exact differential, which sees only its endpoints.
Combining this with the work-energy theorem, $T_2 - T_1 = U_1 - U_2$,
which rearranges to
\[
T_1 + U_1 = T_2 + U_2,
\]
the conservation of mechanical energy. The chain is the whole logical
spine of the energy method: Newton's law dotted with velocity gives the
work-energy theorem; the path-independence of a conservative force turns
its work into a potential difference; the two together conserve $T+U$.

\input{volumes/03-force/ch05-energy-methods/figures/fig-path-independence}

When a non-conservative force is also present, split the total force into
its conservative part $-\nabla U$ and a remainder $\vec{F}^{\text{nc}}$.
The work-energy theorem still holds for the total force, so
\[
T_2 - T_1 = (U_1 - U_2) + \int_1^2 \vec{F}^{\text{nc}}\cdot d\vec{r},
\]
which rearranges to
\[
(T_2 + U_2) - (T_1 + U_1) = \int_1^2 \vec{F}^{\text{nc}}\cdot d\vec{r}
= W^{\text{nc}}.
\]
The change in mechanical energy equals the work done by the
non-conservative forces. For friction $W^{\text{nc}}$ is negative and
the mechanical energy falls; for an applied drive it may be either sign.
The friction worked example later in the chapter is this one equation
applied twice along a round trip, and the rate form $d(T+U)/dt =
\vec{F}^{\text{nc}}\cdot\vec{v}$ is the power balance that the balance
archetype names.

\section{Lagrangian mechanics introduced}\pathtag{core}

A mechanical system with $n$ generalised coordinates $q_1, q_2, \ldots,
q_n$ (independent quantities that locate the system completely, given
the constraints) has a kinetic energy $T = T(q, \dot{q}, t)$ and a
potential energy $U = U(q, t)$. The \engterm{Lagrangian} is
\[
L(q, \dot{q}, t) = T(q, \dot{q}, t) - U(q, t).
\]
The Lagrangian is a scalar function of $2n + 1$ arguments: the $n$
generalised coordinates, the $n$ generalised velocities, and time.

The equations of motion are the \engterm{Euler-Lagrange equations}:
for each generalised coordinate $q_k$, $k = 1, \ldots, n$,
\[
\frac{d}{dt}\left(\frac{\partial L}{\partial \dot{q}_k}\right) -
\frac{\partial L}{\partial q_k} = Q_k^{\text{nc}},
\]
where $Q_k^{\text{nc}}$ is the generalised non-conservative force
conjugate to $q_k$ (zero if all forces are conservative). The
left-hand side is constructed from the Lagrangian in a mechanical way:
take the partial derivative of $L$ with respect to the generalised
velocity, then differentiate the result with respect to time; subtract
the partial derivative of $L$ with respect to the corresponding
generalised coordinate. The result, set equal to the generalised
non-conservative force, is the equation of motion for that coordinate.

The Euler-Lagrange equations are mathematically equivalent to
Newton's second law applied to every particle, with constraint forces
solved for. The equivalence is provable: starting from d'Alembert's
principle and a coordinate transformation from Cartesian particle
positions to generalised coordinates, the inertial term reduces
algebraically to $d/dt(\partial T/\partial \dot{q}) - \partial T/
\partial q$, and the conservative force terms to $-\partial U/\partial
q$. The Lagrangian formulation is therefore not a new physics; it is
a more economical bookkeeping of the same physics.

The economy is the key engineering benefit. A robot arm with $n$
joints has $n$ generalised coordinates (the joint angles); writing
its equations of motion in Newtonian form requires drawing $n + 1$
free-body diagrams (one per link plus the base reactions) and solving
$3n + 3$ equations in the joint reaction forces, then eliminating
$2n$ unknown reaction components. The Lagrangian form writes one
scalar $L$, takes $n$ partial-derivative pairs, and arrives at $n$
equations of motion directly. For $n > 2$ the difference is large.

\engterm{Cyclic coordinates and conserved momenta}. A generalised
coordinate that does not appear in $L$ (only $\dot{q}_k$ appears) is
\engterm{cyclic}. For cyclic $q_k$, the Euler-Lagrange equation
reduces to
\[
\frac{d}{dt}\left(\frac{\partial L}{\partial \dot{q}_k}\right) = 0,
\]
which states that the corresponding generalised momentum $p_k =
\partial L / \partial \dot{q}_k$ is constant. Identifying cyclic
coordinates is the standard route to conservation laws: a system
whose $L$ does not depend on a Cartesian coordinate conserves the
corresponding linear momentum; a system whose $L$ does not depend on
a rotation angle conserves the corresponding angular momentum. The
connection between symmetries of $L$ and conservation laws is the
theorem of Emmy Noether (1918), which extends to continuum mechanics
and field theory in the standard way.

\textbf{Worked example: a bead on a frictionless wire.} The smallest
non-trivial application makes the bookkeeping concrete. A bead of mass
$m$ slides on a straight frictionless horizontal wire with no applied
force. The one generalised coordinate is the position $x$ along the
wire. The kinetic energy is $T = \tfrac12 m \dot{x}^2$, the potential
energy is constant and may be set to zero, so $L = \tfrac12 m \dot{x}^2$.
The coordinate $x$ does not appear in $L$, only $\dot{x}$ does, so $x$
is cyclic and the conjugate momentum $p = \partial L/\partial\dot{x} =
m\dot{x}$ is conserved. The Euler-Lagrange equation confirms it:
$\tfrac{d}{dt}(m\dot{x}) - 0 = 0$, hence $m\ddot{x} = 0$, the bead
moves at constant velocity. The conserved momentum is the linear
momentum, and its conservation is the symmetry of the Lagrangian under
translation along the wire. The same logic, applied to a rotation
coordinate that is absent from $L$, conserves angular momentum; applied
to time, when $L$ has no explicit time dependence, it conserves energy.
Conservation laws are read off the absences in the Lagrangian.

\begin{mastery}
\textbf{The Hamiltonian and phase space.} The Lagrangian formulation
works in the $2n$-dimensional space of coordinates and velocities
$(q, \dot{q})$. The Hamiltonian formulation, due to Hamilton (1834) and
developed by Jacobi, trades each velocity $\dot{q}_k$ for its conjugate
momentum $p_k = \partial L/\partial\dot{q}_k$ and works in the phase
space of $(q, p)$. The exchange is a Legendre transform, the same
construction that carries internal energy to enthalpy in
thermodynamics:
\[
H(q, p, t) = \sum_k p_k \dot{q}_k - L(q, \dot{q}, t),
\]
with the velocities eliminated in favour of the momenta. For the broad
class of systems whose kinetic energy is a homogeneous quadratic in the
velocities and whose potential is velocity-independent, $H = T + U$,
the total energy. The equations of motion become Hamilton's canonical
equations, a symmetric first-order pair,
\[
\dot{q}_k = \frac{\partial H}{\partial p_k},
\qquad
\dot{p}_k = -\frac{\partial H}{\partial q_k}.
\]
The pair has $2n$ first-order equations in place of $n$ second-order
ones, a form that suits numerical integration, that exposes the
geometry of phase space (Liouville's theorem on the conservation of
phase-space volume, the symplectic structure preserved by the flow),
and that is the bridge to statistical mechanics and to the canonical
quantisation of quantum mechanics. The Landau-Lifshitz treatment
develops the canonical formalism from the action principle with minimal
apparatus \cite{text:landau-mechanics}. For the working analysis in
this chapter the Lagrangian form is sufficient; the Hamiltonian form is
the entry point to the deeper structure and to the symplectic
integrators used when long-time energy conservation matters.
\end{mastery}

\section{Generalised coordinates and constraints}\pathtag{standard}

A mechanical system in three-dimensional space with $N$ particles
has $3N$ Cartesian coordinates. Constraints reduce the system's
freedom. A \engterm{holonomic} constraint is an equation of the form
\[
f(r_1, r_2, \ldots, r_N, t) = 0,
\]
relating the particle positions and time. Each independent holonomic
constraint reduces the degrees of freedom by one. A system with $N$
particles and $c$ holonomic constraints has $n = 3N - c$ degrees of
freedom, and $n$ independent generalised coordinates suffice to
locate it.

Examples of holonomic constraints. A particle confined to a surface
$f(x, y, z) = 0$ has one constraint, two degrees of freedom; the
surface coordinates are the natural generalised coordinates. A bead
on a rotating circular hoop has the constraint $x^2 + y^2 = R^2$
combined with the rotating-frame angle; one generalised coordinate
(the bead's angular position on the hoop) suffices. A double pendulum
in two dimensions has four Cartesian coordinates (two per bob) and
two constraints (each bob's distance from its pivot is constant);
two generalised coordinates (the two swing angles) suffice.

A \engterm{nonholonomic} constraint involves the velocities and
cannot be integrated into a relation among the coordinates alone. The
canonical example is a wheel rolling without slipping on a surface:
the rolling condition $\vec{v}_{\text{contact}} = 0$ relates the
wheel's velocity to its angular velocity, but the integrated form is
a path-dependent relation that does not reduce to a constraint on the
coordinates. Nonholonomic systems retain more generalised
coordinates than their kinematic degrees of freedom suggest; the
analysis uses Lagrange multipliers or the Gibbs-Appell formulation,
both beyond this chapter's scope. The robotics and vehicle-dynamics
treatments in Volume VIII (chapter 8) develop the nonholonomic case.

The choice of generalised coordinates is the engineer's. Independent
coordinates that completely locate the system are required; beyond
that, the choice is tactical. A pendulum's swing angle $\theta$ is
the natural coordinate; the bob's Cartesian position is allowed but
introduces a constraint $x^2 + y^2 = L^2$ that the swing angle
absorbs automatically. A system whose constraints are mismatched
with its coordinate choice produces algebraic complication; matched
coordinates produce algebraic simplicity. The standard heuristic is
that coordinates should be chosen so that each constraint is
identically satisfied by the chosen coordinate set.

A short catalogue fixes the count for the systems the chapter uses. A
single particle free in space has three degrees of freedom; confined to a
surface, two; confined to a curve, one. A rigid body free in space has six,
three of translation and three of rotation; a rigid body in a plane has
three, two of translation and one of rotation. A planar mechanism of $b$
rigid bodies connected by $j$ pin joints, each pin removing two of the
relative degrees of freedom, has $3b - 2j$ degrees of freedom by the
Gr\"ubler count, the formula that tells a linkage designer whether a
mechanism moves with one input ($3b - 2j = 1$), is a rigid structure
($3b - 2j = 0$), or is a mechanism with redundant freedom ($3b - 2j > 1$).
A four-bar linkage, four links (counting the fixed ground link) and four
pins, scores $3\cdot 4 - 2\cdot 4 = 4 \ne$ the naive count because the
ground link is fixed; counting only the three moving links gives $3\cdot 3 -
2\cdot 4 = 1$, the single degree of freedom that makes the four-bar the
workhorse of constrained motion. The same count applied to a structure that
should not move is the determinacy check: a planar truss of $b$ bars, $j$
joints, and $r$ support reactions is determinate when $b + r = 2j$,
indeterminate when $b + r > 2j$ (the redundant-structure case the least-work
method resolves), and a mechanism when $b + r < 2j$.

\engterm{Degrees of freedom}, the count of independent generalised
coordinates required, is the most important number to determine
correctly before any energy-method analysis. A wrong count produces
the failure mode of section 5.7: an analysis whose equations of
motion miss a mode of the real system.

\section{Hamilton's principle}\pathtag{standard}

Among all conceivable trajectories $q(t)$ from an initial state $q
(t_1) = q_1$ to a final state $q(t_2) = q_2$ at fixed times $t_1$
and $t_2$, the trajectory the system actually follows is the one
that makes the \engterm{action}
\[
S = \int_{t_1}^{t_2} L(q, \dot{q}, t) \,dt
\]
stationary under arbitrary variations of $q(t)$ with the endpoints
fixed. This is \engterm{Hamilton's principle}, established by William
Rowan Hamilton in his 1834 paper ``On a General Method in Dynamics.''
Figure~\ref{fig:vol03ch05:action-paths} draws the construction: among
all admissible trial paths sharing the two endpoints, the physical path
is the one whose action does not change to first order under a small
deformation.

\input{volumes/03-force/ch05-energy-methods/figures/fig-action-paths}

The Euler-Lagrange equations are the necessary conditions for $S$ to
be stationary. Varying the action with respect to a small perturbation
$\delta q(t)$ that vanishes at the endpoints,
\[
\delta S = \int_{t_1}^{t_2} \left[\frac{\partial L}{\partial q}
\delta q + \frac{\partial L}{\partial \dot{q}} \delta \dot{q}\right]
\,dt.
\]
Integrating the second term by parts (the boundary terms vanish
because $\delta q$ vanishes at the endpoints):
\[
\delta S = \int_{t_1}^{t_2} \left[\frac{\partial L}{\partial q} -
\frac{d}{dt}\frac{\partial L}{\partial \dot{q}}\right] \delta q
\,dt.
\]
For $\delta S = 0$ to hold for arbitrary $\delta q(t)$, the bracketed
expression must vanish at every $t$:
\[
\frac{d}{dt}\frac{\partial L}{\partial \dot{q}} - \frac{\partial L}{
\partial q} = 0,
\]
the Euler-Lagrange equation. The same derivation in $n$ coordinates
gives the $n$ Euler-Lagrange equations of section 5.3.

The variational form is more than a derivation route. It expresses
mechanics as an optimisation: the trajectory the system follows is
optimal among all conceivable trajectories, with respect to a
specific functional (the action integral). The optimisation
interpretation generalises beyond classical mechanics. Hamilton's
principle is the starting point for the modern formulation of
continuum mechanics, fluid dynamics, electromagnetism, general
relativity, and quantum field theory; the action principle is the
single most general way to specify a physical theory.

The engineering payoff is, again, economy. Once the system's
Lagrangian is identified, the equations of motion follow mechanically.
The variational form also gives a route to approximate solutions: an
assumed trial trajectory $q(t; a_1, a_2, \ldots, a_m)$ with $m$
adjustable parameters $a_k$ produces an action $S(a_1, a_2, \ldots,
a_m)$ that is a function (not a functional) of the $m$ parameters;
making $S$ stationary with respect to the parameters yields an
approximate equation of motion that is exact within the trial family
and is the basis of the Rayleigh-Ritz method, the finite-element
method, and most numerical solvers in structural dynamics.

\begin{mastery}
\textbf{The Rayleigh quotient as an energy estimate.} The variational
structure gives a fast, conservative estimate of a vibrating system's
lowest natural frequency without solving the equations of motion. For a
linear conservative system with a trial mode shape, the ratio of peak
potential energy to a reference kinetic energy bounds the fundamental
frequency from above. In the continuous case of a beam or string, the
Rayleigh quotient is
\[
\omega^2 \le R[\phi]
= \frac{\displaystyle\int (\text{stiffness})\,(\phi')^2\,dx}
       {\displaystyle\int (\text{mass density})\,\phi^2\,dx},
\]
evaluated on any admissible trial shape $\phi$ that satisfies the
geometric boundary conditions. The estimate is always an over-estimate
of the true fundamental, and it is remarkably accurate even for crude
trial shapes, because the quotient is stationary at the true mode and
so is insensitive to first-order error in $\phi$. A parabolic trial
shape for a cantilever, for instance, lands within a few per cent of
the exact fundamental frequency. The method is the practising
structural engineer's back-of-envelope check on a finite-element
natural-frequency result: if the code reports a fundamental far from the
Rayleigh estimate, the model, not the structure, is usually at fault.
The same quotient, generalised to a matrix eigenvalue problem,
underlies the power-iteration and subspace methods that the
finite-element solvers themselves use. Den~Hartog's vibrations text
develops the Rayleigh method as the working tool for hand estimates of
natural frequencies \cite{text:den-hartog-vibrations}.
\end{mastery}

\begin{archetype}[Optimisation]
Variational principles cast mechanics as optimisation: among
admissible trajectories, the physical one extremises a scalar
functional. The same archetype recurs in thermodynamics (equilibrium
minimises Gibbs free energy), in chemistry (reactions follow
free-energy gradients), in optics (Fermat's principle: light follows
the path of stationary optical length), in elasticity (the deformed
state minimises elastic potential energy), in finite-element
analysis (the discretised approximation minimises a discrete energy
functional), and in machine learning (training minimises a loss
function). The optimisation archetype is one of the unifying
patterns of engineering mathematics; chapter 4 of Volume IX develops
it as a stand-alone framework.
\end{archetype}

\section{Worked examples}\pathtag{standard}

\subsection{Simple pendulum}

A simple pendulum consists of a bob of mass $m$ suspended from a
fixed pivot by a massless inextensible string of length $L$. The
generalised coordinate is the swing angle $\theta$, measured from
the downward vertical.

\textbf{Kinetic energy.} The bob's velocity is $L\dot{\theta}$
tangent to its circular path, so $T = \tfrac{1}{2} m L^2 \dot{\theta}^2$.

\textbf{Potential energy.} With $U = 0$ at the lowest point of the
swing, the bob's height above the lowest point is $L(1 - \cos\theta)$,
so $U = m g L (1 - \cos\theta)$.

\textbf{Lagrangian.}
\[
L = T - U = \tfrac{1}{2} m L^2 \dot{\theta}^2 - m g L (1 - \cos\theta).
\]

\textbf{Euler-Lagrange equation.}
\[
\frac{\partial L}{\partial \dot{\theta}} = m L^2 \dot{\theta}, \quad
\frac{d}{dt}\frac{\partial L}{\partial \dot{\theta}} = m L^2
\ddot{\theta}, \quad \frac{\partial L}{\partial \theta} = -m g L
\sin\theta.
\]
Therefore $m L^2 \ddot{\theta} + m g L \sin\theta = 0$, or
\[
\ddot{\theta} + \frac{g}{L}\sin\theta = 0.
\]
This is the equation of motion of a simple pendulum, identical to
the result of a Newtonian analysis with free-body diagrams of the
bob and constraint reactions at the pivot. The Lagrangian derivation
involved one kinetic-energy scalar, one potential-energy scalar, and
mechanical differentiation; the Newtonian derivation requires the
free-body diagram, the resolution of the tension in the string into
radial and tangential components, and the explicit elimination of
the unknown tension to leave the tangential equation.

For small amplitudes ($\theta \ll 1$), $\sin\theta \approx \theta$ and
the equation linearises to $\ddot{\theta} + (g/L)\theta = 0$, the
simple harmonic oscillator with period $T = 2\pi\sqrt{L/g}$. A
$1\,\si{\meter}$ pendulum has period approximately $2.0\,\si{\second}$.

At larger amplitude the small-angle period under-predicts the truth,
because the restoring torque $-mgL\sin\theta$ grows more slowly than the
linear law $-mgL\,\theta$ that the small-angle formula assumes. The
leading correction is $T \approx T_0\,(1 + \theta_0^2/16)$, and the
exact period is the complete elliptic integral of the first kind. The
script \path{code/pendulum_period.py} integrates the full nonlinear
equation, measures the period by detecting the return to the turning
point, and writes the comparison to \path{data/pendulum_period.csv}: at
$\theta_0 = 45^\circ$ the small-angle formula is low by about
$4\,\%$ and the first correction is good to better than $0.2\,\%$;
at $\theta_0 = 90^\circ$ the small-angle formula is low by $18\,\%$
and even the first correction is short by $2\,\%$. The exercises return
to these numbers.

\subsection{Double pendulum}

A double pendulum has two bobs of masses $m_1$ and $m_2$ on
inextensible massless strings of lengths $L_1$ and $L_2$, the second
suspended from the first. Generalised coordinates: the swing angles
$\theta_1$ (upper) and $\theta_2$ (lower), both measured from the
downward vertical (figure~\ref{fig:vol03ch05:double-pendulum}).

\input{volumes/03-force/ch05-energy-methods/figures/fig-double-pendulum}

\textbf{Cartesian positions of the bobs.}
\begin{align*}
(x_1, y_1) &= (L_1 \sin\theta_1, -L_1 \cos\theta_1), \\
(x_2, y_2) &= (L_1 \sin\theta_1 + L_2 \sin\theta_2, -L_1 \cos\theta_1
- L_2 \cos\theta_2).
\end{align*}

\textbf{Kinetic energy.} Differentiating the positions and computing
$\tfrac{1}{2}m_1(\dot{x}_1^2 + \dot{y}_1^2) + \tfrac{1}{2}m_2(\dot{x}_2^2
+ \dot{y}_2^2)$ gives, after simplification,
\[
T = \tfrac{1}{2}(m_1 + m_2) L_1^2 \dot{\theta}_1^2 + \tfrac{1}{2}
m_2 L_2^2 \dot{\theta}_2^2 + m_2 L_1 L_2 \dot{\theta}_1 \dot{\theta}_2
\cos(\theta_1 - \theta_2).
\]

\textbf{Potential energy.}
\[
U = -(m_1 + m_2) g L_1 \cos\theta_1 - m_2 g L_2 \cos\theta_2 + \text{
const}.
\]

\textbf{Euler-Lagrange equations.} With $L = T - U$, the two
equations are
\begin{align*}
& (m_1 + m_2) L_1^2 \ddot{\theta}_1 + m_2 L_1 L_2 \ddot{\theta}_2 \cos
(\theta_1 - \theta_2) + m_2 L_1 L_2 \dot{\theta}_2^2 \sin(\theta_1 -
\theta_2) \\
& \quad + (m_1 + m_2) g L_1 \sin\theta_1 = 0, \\
& m_2 L_2^2 \ddot{\theta}_2 + m_2 L_1 L_2 \ddot{\theta}_1 \cos(
\theta_1 - \theta_2) - m_2 L_1 L_2 \dot{\theta}_1^2 \sin(\theta_1 -
\theta_2) \\
& \quad + m_2 g L_2 \sin\theta_2 = 0.
\end{align*}

The pair is coupled, nonlinear, and analytically intractable except
in special cases (small amplitudes, in-phase motion, the limit
$m_1 \to \infty$ or $m_1 \to 0$). The general motion is chaotic in
the technical sense: the system is deterministic but exhibits
exponential sensitivity to initial conditions. The double pendulum
is the simplest mechanical demonstration of deterministic chaos and
is the project of this chapter.

The Newtonian derivation of these same equations requires drawing two
free-body diagrams, identifying the two unknown tensions (one in each
string), resolving each tension into Cartesian or polar components,
and eliminating the tensions from the coupled equations of motion. The
algebra is several pages and prone to sign errors. The Lagrangian
derivation is the two-page computation above, with no constraint
forces to eliminate.

\textbf{Small-amplitude normal modes.} For equal masses $m_1 = m_2 = m$
and equal lengths $L_1 = L_2 = L$, the small-amplitude limit
($\theta_i \ll 1$, $\cos(\theta_1-\theta_2)\to 1$, $\sin\theta_i\to
\theta_i$, products of velocities dropped) linearises the pair to
\[
2\ddot{\theta}_1 + \ddot{\theta}_2 + 2\,\tfrac{g}{L}\,\theta_1 = 0,
\qquad
\ddot{\theta}_1 + \ddot{\theta}_2 + \tfrac{g}{L}\,\theta_2 = 0.
\]
Seeking solutions $\theta_i = a_i \cos\omega t$ turns this into the
generalised eigenvalue problem $(K - \omega^2 M)\,a = 0$, with
\[
M = \begin{pmatrix} 2 & 1 \\ 1 & 1 \end{pmatrix},
\qquad
K = \frac{g}{L}\begin{pmatrix} 2 & 0 \\ 0 & 1 \end{pmatrix}.
\]
The two roots are $\omega^2 = (2 \mp \sqrt{2})\,g/L$, that is
\[
\omega_1^2 = (2 - \sqrt2)\,\tfrac{g}{L} \approx 0.586\,\tfrac{g}{L},
\qquad
\omega_2^2 = (2 + \sqrt2)\,\tfrac{g}{L} \approx 3.414\,\tfrac{g}{L}.
\]
The low-frequency mode has the two bobs swinging in phase
($a_2/a_1 = +\sqrt2/2 \approx 0.707$); the high-frequency mode has them
out of phase ($a_2/a_1 = -\sqrt2 \approx -1.414$). The file
\path{data/double_pendulum_modes.csv} carries the two modes' frequency
factors and amplitude ratios. The double pendulum is thus orderly at
small amplitude, a pair of clean normal modes, and chaotic at large
amplitude, where the linearisation that produced the modes no longer
holds.

\textbf{Numerical exploration and sensitivity.} The full nonlinear pair
has no closed-form solution and is integrated numerically. The
script \path{code/double_pendulum.py} solves the equations of motion
with a fixed-step fourth-order Runge-Kutta integrator, inverting the
two-by-two mass matrix at each step to extract the angular
accelerations, and logs the total mechanical energy as the
accuracy check: a correct conservative integration holds the energy
constant to within the integrator's truncation error, and the run
recorded in \path{data/double_pendulum_trajectory.csv} drifts by under
$2\times 10^{-7}$ in relative terms over thirty seconds. Running the
same initial condition twice with the upper angle perturbed by
$10^{-6}\,\si{\radian}$ produces trajectories whose separation grows
exponentially until it saturates at order one radian, the signature of
deterministic chaos plotted in
figure~\ref{fig:vol03ch05:double-pendulum-divergence}. The equations
are exact and the system is deterministic; it is also unpredictable in
practice past a few seconds, which is the distinction between
determinism and predictability that the chaos literature makes
central.

\input{volumes/03-force/ch05-energy-methods/figures/fig-double-pendulum-divergence}

\subsection{Bead on a rotating hoop}

A bead of mass $m$ slides without friction on a vertical circular
hoop of radius $R$ that rotates about its vertical diameter with
constant angular velocity $\Omega$. Find the bead's stable
equilibrium positions on the hoop as a function of $\Omega$.

\textbf{Generalised coordinate.} The angle $\theta$ measuring the
bead's position on the hoop from the bottom; $\theta = 0$ is the
lowest point, $\theta = \pi$ the highest.

\textbf{Kinetic energy.} The bead's velocity has two components: a
tangential velocity $R\dot{\theta}$ along the hoop, and a tangential
velocity $R\sin\theta \cdot \Omega$ due to the hoop's rotation. The
two components are perpendicular, so
\[
T = \tfrac{1}{2} m (R\dot{\theta})^2 + \tfrac{1}{2} m (R\sin\theta
\cdot \Omega)^2 = \tfrac{1}{2} m R^2 (\dot{\theta}^2 + \Omega^2
\sin^2\theta).
\]

\textbf{Potential energy.} The bead's height above the bottom is
$R(1 - \cos\theta)$, so $U = m g R (1 - \cos\theta)$.

\textbf{Lagrangian and equation of motion.} The Lagrangian $L = T - U$
gives
\[
m R^2 \ddot{\theta} - m R^2 \Omega^2 \sin\theta \cos\theta + m g R
\sin\theta = 0.
\]
At equilibrium $\ddot{\theta} = 0$, so
\[
\sin\theta \,(g - R \Omega^2 \cos\theta) = 0.
\]
The solutions are $\theta = 0$ (bottom), $\theta = \pi$ (top), and,
when $R\Omega^2 \cos\theta = g$ has a solution in $(0, \pi/2)$, two
additional equilibria $\theta = \pm \theta_c$ with $\cos\theta_c =
g/(R\Omega^2)$.

The additional equilibria exist only when $R\Omega^2 > g$, i.e.,
when the centripetal acceleration at the equator of the hoop exceeds
gravity. Below the threshold, the bead's stable position is at the
bottom; above the threshold, the bottom becomes unstable and the
bead settles at $\pm \theta_c$. The transition at $R\Omega^2 = g$ is
a \engterm{bifurcation}, the simplest mechanical example of a
symmetry-breaking transition; it is the prototype for many phase
transitions in continuous physical systems. The branch diagram in
figure~\ref{fig:vol03ch05:hoop-bifurcation} draws the pitchfork: the
bottom equilibrium loses stability at the threshold and two symmetric
off-axis equilibria appear and take it over.

The stability of each equilibrium is read directly off an effective
potential. Collecting the position-dependent terms of the Lagrangian
into $U_\text{eff}(\theta) = m g R(1 - \cos\theta) - \tfrac12 m R^2
\Omega^2 \sin^2\theta$, the equilibria are its stationary points and the
stable ones are its minima, identified by the sign of $U_\text{eff}''$.
Below the threshold the bottom is the sole minimum; above it the bottom
becomes a maximum and the two off-axis points become the minima. The
rotation term has reshaped the potential, lowering the energy of
off-axis positions until they out-compete the bottom. The script
\path{code/hoop_bifurcation.py} tabulates the equilibrium angles and
their stability across the threshold into
\path{data/hoop_equilibria.csv}, and the symbolic derivation of the
equation of motion is in \path{code/lagrangian_symbolic.py}, which
generates the Euler-Lagrange equation for this system and for the
simple pendulum from the Lagrangian alone, with no hand algebra.

\input{volumes/03-force/ch05-energy-methods/figures/fig-hoop-bifurcation}

\section{Strain energy and the deflection of structures}\pathtag{standard}

The dynamics of sections 5.3 through 5.6 use kinetic and potential
energy of moving masses. A second, equally large branch of energy
methods works on structures at rest: a loaded elastic member stores
\engterm{strain energy} as it deforms, and the deflection of the
structure under load can be recovered from that scalar by
differentiation, with no need to integrate the deflected shape or solve
the governing differential equation. The two branches share the same
discipline. Write one scalar energy, differentiate it in a standard
pattern, and read the quantity of interest off the result.

\engterm{Strain energy} is the elastic potential energy stored in a
deformed member, equal to the work done by the loads in deforming it.
For an axially loaded bar of length $L$, cross-sectional area $A$, and
modulus $E$ carrying axial force $N$, the stored energy is
\[
U = \frac{N^2 L}{2 E A},
\]
the area under the linear force-extension line, $\tfrac12 N \delta$ with
$\delta = NL/(EA)$. For a beam in bending with internal bending moment
$M(x)$, second moment of area $I$, and modulus $E$, the strain energy is
the integral of the bending contribution along the member,
\[
U = \int_0^L \frac{M(x)^2}{2 E I}\,dx,
\]
with parallel expressions for torsion ($T^2 L / (2GJ)$ per uniform shaft)
and for transverse shear (usually small in slender members and dropped
here). Each is a quadratic, positive form in the internal load, the
structural analogue of the $\tfrac12 k x^2$ spring well of
section 5.1.

The four forms (axial, bending, torsion, shear) share a single template:
the strain energy of a member is the integral along its length of the square
of an internal load over twice a stiffness, $\int (\text{load})^2 /
(2\cdot\text{stiffness})\,dx$. The stiffness in the denominator is the
member's resistance to the deformation the load drives, $EA$ for stretch,
$EI$ for bending curvature, $GJ$ for twist, $GA_s$ for shear slip. A member
carrying several load types at once stores the sum of the four energies,
since the deformations are independent to first order, and the analysis adds
them without cross terms. The recurring economy of the static branch matches
the dynamic one: where the Lagrangian replaced a free-body diagram per body
with two scalar energies, the strain-energy method replaces a deflected-shape
integration or a differential-equation solve with one scalar energy whose
derivative is the wanted displacement. The two branches are the same
discipline applied to motion and to deformation, and the worked examples
that follow move between them, a deflection here, a redundant reaction there,
an impact force after that, all read off a single scalar by differentiation
or stationarity.

\subsection{Strain energy from the stress field}

The member-level formulas above are not postulates; each is the volume
integral of a single energy density, and deriving them from that density
fixes their structure and shows where the shear correction lives. A small
element of an elastic solid carrying a uniaxial stress $\sigma$ strains by
$\varepsilon = \sigma/E$, and as the stress is built up from zero the
element stores work per unit volume equal to the triangular area under the
linear stress-strain line,
\[
u = \int_0^{\varepsilon}\sigma\,d\varepsilon
= \int_0^{\varepsilon}E\varepsilon'\,d\varepsilon'
= \tfrac12 E\varepsilon^2
= \frac{\sigma^2}{2E},
\]
the \engterm{strain energy density}, drawn in
figure~\ref{fig:vol03ch05:strain-energy-density}. A shear stress $\tau$
producing shear strain $\gamma = \tau/G$ stores $u = \tau^2/(2G)$ by the
identical triangle. The total strain energy of a member is the integral of
$u$ over its volume, and the four member formulas follow by writing the
stress in terms of the internal load and integrating.

\input{volumes/03-force/ch05-energy-methods/figures/fig-strain-energy-density}

\textbf{Axial.} A bar carrying axial force $N$ has the uniform stress
$\sigma = N/A$ over its cross-section, so $u = N^2/(2EA^2)$ and the volume
integral over a length $L$ of constant area $A$ is
\[
U = \int_V u\,dV = \int_0^L \frac{N^2}{2EA^2}\,A\,dx
= \frac{N^2 L}{2EA},
\]
the axial formula, with the area in the denominator because a fatter bar
spreads the same force over more material and stores less energy per unit
length.

\textbf{Bending.} A beam in bending carries the linear stress distribution
$\sigma(y) = M y/I$, with $y$ measured from the neutral axis. The density
$u = \sigma^2/(2E) = M^2 y^2/(2EI^2)$ integrates over the cross-section
through the definition of the second moment of area, $\int_A y^2\,dA = I$,
so the energy per unit length is $M^2 I/(2EI^2) = M^2/(2EI)$ and the energy
along the member is
\[
U = \int_0^L \frac{M(x)^2}{2EI}\,dx,
\]
the bending formula, with the moment squared because energy cannot care
about the sign of the curvature.

\textbf{Torsion.} A circular shaft in torsion carries the shear stress
$\tau(r) = T r/J$ growing linearly from the axis. The density $\tau^2/(2G)
= T^2 r^2/(2GJ^2)$ integrates over the section through $\int_A r^2\,dA = J$
to give $T^2/(2GJ)$ per unit length, hence $U = T^2 L/(2GJ)$ for a uniform
shaft, the torsion formula the stepped-shaft example uses.

\textbf{Shear.} Transverse shear stores $u = \tau^2/(2G)$ from the parabolic
shear-stress distribution across the section; integrating and writing the
result with an effective shear area $A_s$ gives $U_{\text{shear}} = \int
V^2/(2GA_s)\,dx$ with $V$ the shear force. For a slender member the bending
energy dominates the shear energy by a factor of order $(L/h)^2$ with $h$ the
section depth, which is why the shear term is dropped throughout this chapter
and retained only for deep or short members. The four densities are one idea,
the triangle $\tfrac12\times\text{stress}\times\text{strain}$, integrated over
the volume with the stress written from the internal load.

\subsection{Castigliano's theorems and minimum potential energy derived}

The deflection results of this section all descend from one fact about a
linearly elastic structure: its strain energy is a state function of the
loads (equivalently of the displacements), so derivatives of that scalar
return the conjugate kinematic or static quantities. We derive the four
working statements in turn.

\textbf{Castigliano's first theorem} reads a load off the strain energy as a
function of displacements. Apply a set of loads $P_i$ through their work-
conjugate displacements $\delta_i$, and write the strain energy
$U(\delta_1,\delta_2,\ldots)$ as a function of those displacements. Give one
displacement a small increment $d\delta_i$ at fixed others. The work the
loads do is $P_i\,d\delta_i$ to first order, and for an elastic body that work
is stored as the energy increment $dU = (\partial U/\partial\delta_i)\,
d\delta_i$. Equating the two,
\[
P_i = \frac{\partial U}{\partial\delta_i},
\]
the load as the derivative of strain energy with respect to its conjugate
displacement. This is the static sibling of the Lagrangian result
$\partial L/\partial q$, a force read off a scalar energy.

\textbf{Castigliano's second theorem} reads a displacement off the
\emph{complementary} energy as a function of loads. For a linear structure
the force-deflection line is straight, so the strain energy and the
complementary energy are numerically equal, $U = U^*$, and the theorem is
stated against $U$ directly. Increment one load by $dP_i$ at fixed others.
The complementary work the structure absorbs is $\delta_i\,dP_i$, equal to
the energy increment $(\partial U/\partial P_i)\,dP_i$, so
\[
\delta_i = \frac{\partial U}{\partial P_i},
\qquad
\theta_i = \frac{\partial U}{\partial M_i},
\]
a deflection conjugate to a force, or a rotation conjugate to a couple, as
the derivative of strain energy with respect to that load. For a member whose
energy is $\int M^2/(2EI)\,dx$ the derivative passes inside the integral,
\[
\delta_i = \frac{\partial}{\partial P_i}\int \frac{M^2}{2EI}\,dx
= \int \frac{M}{EI}\,\frac{\partial M}{\partial P_i}\,dx,
\]
which is the form the frame, curved-member, and propped-cantilever examples
evaluate.

\textbf{The unit-load (dummy-load) method} is the same integral with the
load derivative read as a virtual force. The factor $\partial M/\partial P_i$
is the bending moment per unit of the load $P_i$, that is the moment $m$ that
a unit load applied in the place and direction of the wanted deflection would
produce. Writing $m = \partial M/\partial P_i$ turns Castigliano's integral
into
\[
\delta = \int \frac{M\,m}{EI}\,dx,
\qquad
\delta = \sum_i \frac{N_i\,n_i\,L_i}{E_i A_i}
\]
for beams and for pin-jointed trusses respectively, with $M,N_i$ the real
internal loads and $m,n_i$ the internal loads under the virtual unit load.
The method derives equally from virtual work: load the structure with the
real system, then apply a virtual unit load and let the structure move
through its real displacements; the external virtual work $1\cdot\delta$
equals the internal virtual work $\int m\,(M/EI)\,dx$, the virtual moments
moving through the real curvatures $M/EI$. The two derivations meet because
both rest on the same state-function energy. The dummy-load trick for a
deflection in a direction carrying no real load is now transparent: apply a
fictitious load $Q$ there, carry it through $M$, differentiate, and set
$Q=0$, since $\partial M/\partial Q$ at $Q=0$ is exactly the unit-load
moment.

\textbf{The principle of minimum total potential energy} is the
displacement-based variational statement that the spring example and the
snap-through arch use. Define the total potential energy of a loaded elastic
structure as its stored strain energy minus the work the applied loads have
done,
\[
\Pi = U - \sum_i P_i\,\delta_i.
\]
At equilibrium $\Pi$ is stationary under any kinematically admissible
variation of the displacement field. The stationarity condition recovers the
equilibrium equations: varying $\delta_i$,
\[
\frac{\partial\Pi}{\partial\delta_i}
= \frac{\partial U}{\partial\delta_i} - P_i = 0
\quad\Longrightarrow\quad
P_i = \frac{\partial U}{\partial\delta_i},
\]
which is Castigliano's first theorem again, now read as the equilibrium that
makes $\Pi$ stationary. The stationary point is a \emph{minimum} when the
equilibrium is stable, the curvature $\partial^2\Pi/\partial\delta_i^2 > 0$
playing the role the spring stiffness $k$ played in the single-coordinate
well of section 5.1. Stationarity gives the equilibrium; the sign of the
second variation gives its stability, the pairing the mastery box on the
second variation develops in full.

\begin{mastery}
\textbf{Clapeyron's theorem and the two-sided energy bound.} A single
equality and a pair of inequalities organise the whole strain-energy
machinery. The equality is \engterm{Clapeyron's theorem}: for a linearly
elastic structure loaded gradually from rest, the strain energy stored
equals half the work the final loads would do moving through the final
displacements,
\[
U = \tfrac12\sum_i P_i\,\delta_i,
\]
the factor of one half because the loads and displacements rise together
along a straight line, so the work is the triangle $\tfrac12 P\delta$ and
not the rectangle $P\delta$. The theorem is the structural reading of the
spring's $\tfrac12 k x^2$ and the reason a suddenly applied load, which
arrives at full value before the structure has moved, does twice the work
and so doubles the response, the $h=0$ limit of the impact factor.

The inequalities bracket the true energy from both sides and turn the two
variational principles into a numerical error bar on any approximate
analysis. The principle of minimum total potential energy says that the
true displacement field minimises $\Pi = U - \sum P_i\delta_i$ over all
kinematically admissible fields, so any assumed displacement field, one
that satisfies the geometric boundary conditions but not necessarily
equilibrium, gives a $\Pi$ no lower than the true minimum, and the
corresponding stiffness estimate is too high: a structure constrained to an
assumed shape is artificially stiff. This is why every Rayleigh-Ritz
frequency and every buckling load in this chapter came out as an
over-estimate. The dual principle of minimum complementary energy says that
the true stress field minimises the complementary energy $\Pi^*$ over all
statically admissible stress fields, those that satisfy equilibrium but not
necessarily compatibility, so any assumed equilibrium stress field gives a
stiffness estimate that is too low. The true energy is caught between the
two,
\[
(\text{compatible-stress estimate}) \;\le\;
U_{\text{true}} \;\le\;
(\text{admissible-displacement estimate}),
\]
and an analyst who computes both an upper bound from an assumed displacement
and a lower bound from an assumed stress brackets the exact stiffness without
ever solving the governing equations. The two bounds are the variational
foundation of the displacement (stiffness) and force (flexibility)
formulations of the finite-element method, which inherit the same one-sided
errors: a standard displacement-based element mesh is always stiffer than the
true structure and converges to the exact response from above as the mesh is
refined, a fact a designer uses to know which way a coarse-mesh result is
wrong.
\end{mastery}

\engterm{Castigliano's second theorem}, $\delta_i = \partial U/\partial
P_i$, derived above, is the workhorse of the deflection examples that
follow. The theorem is the static, structural sibling of the
cyclic-coordinate result of section 5.3: a derivative of a scalar energy
yields a kinematic quantity. Carlo Alberto Castigliano stated it in his
1879 \emph{Th\'eorie de l'\'equilibre des syst\`emes \'elastiques}; the
working derivation and a catalogue of applications are in any
mechanical-design reference \cite[ch.~4]{text:shigley-mechanical-design}.

\subsection{End-to-end case study: tip deflection of a bent cantilever
frame}

A complete energy-method analysis, carried from geometry to a checked
number, fixes the procedure better than a statement of the theorem. The
structure is the L-shaped cantilever frame of
figure~\ref{fig:vol03ch05:castigliano-frame}: a horizontal member of
length $a$ joined at a rigid corner to a vertical member of length $b$
built into a wall, with a vertical load $P$ at the free tip. Both members
are steel tubes of the same uniform bending stiffness $EI$. We want the
vertical deflection of the loaded tip.

\input{volumes/03-force/ch05-energy-methods/figures/fig-castigliano-frame}

\textbf{Step 1: internal bending moments.} Walk the structure from the
free end inward, since the load and geometry are simplest there. On the
horizontal member, measure $x$ from the tip toward the corner. The only
load outboard of the cut is $P$ at lever arm $x$, so the bending moment
is
\[
M_1(x) = P x,\qquad 0 \le x \le a.
\]
On the vertical member, measure $y$ from the corner up toward the wall.
Every point on this member sees the full load $P$ acting at the constant
horizontal lever arm $a$ (the horizontal distance from the member to the
tip does not change as $y$ increases), so
\[
M_2(y) = P a,\qquad 0 \le y \le b,
\]
a constant. The axial force in each member (the horizontal member carries
no axial load from a vertical tip force; the vertical member carries the
direct compression $P$) contributes negligibly to the deflection of a
slender frame and we drop it, returning to it as a check at the end.

\textbf{Step 2: total strain energy.} Sum the bending contributions of
the two members,
\[
U = \int_0^a \frac{M_1(x)^2}{2EI}\,dx
  + \int_0^b \frac{M_2(y)^2}{2EI}\,dy
  = \int_0^a \frac{(Px)^2}{2EI}\,dx
  + \int_0^b \frac{(Pa)^2}{2EI}\,dy.
\]
The first integral is $\dfrac{P^2}{2EI}\cdot\dfrac{a^3}{3}$; the second,
with a constant integrand, is $\dfrac{P^2 a^2}{2EI}\cdot b$. So
\[
U = \frac{P^2 a^3}{6EI} + \frac{P^2 a^2 b}{2EI}
  = \frac{P^2 a^2}{2EI}\left(\frac{a}{3} + b\right).
\]

\textbf{Step 3: differentiate for the deflection.} The tip deflection in
the direction of $P$ is Castigliano's derivative,
\[
\delta_\text{tip} = \frac{\partial U}{\partial P}
  = \frac{P a^2}{EI}\left(\frac{a}{3} + b\right)
  = \frac{P a^3}{3EI} + \frac{P a^2 b}{EI}.
\]
The two terms have a clean reading. The first is the deflection of a
plain cantilever of length $a$ under a tip load, $Pa^3/(3EI)$, the
standard result. The second is the extra tip drop caused by the vertical
member bending under the constant moment $Pa$; it scales with $b$ because
a longer vertical member rotates the corner through a larger angle, and
that rotation throws the tip down by the moment arm $a$.

\textbf{Step 4: a number.} Take $a = 1.2\,\si{\meter}$,
$b = 0.8\,\si{\meter}$, a load $P = 2.0\,\si{\kilo\newton}$, and steel
tubing with $E = 200\,\si{\giga\pascal}$ and
$I = 4.0\times 10^{-6}\,\si{\meter\tothe{4}}$ (a $60\,\si{\milli\meter}$
square tube of $\sim 4\,\si{\milli\meter}$ wall, say), so
$EI = 8.0\times 10^{5}\,\si{\newton\meter\squared}$. Then
\[
\delta_\text{tip}
= \frac{2000 \cdot 1.2^2}{8.0\times 10^{5}}
  \left(\frac{1.2}{3} + 0.8\right)
= \frac{2000 \cdot 1.44}{8.0\times 10^{5}} \cdot 1.2
\approx 4.3\times 10^{-3}\,\si{\meter},
\]
a tip deflection of about $4.3\,\si{\milli\meter}$. The cantilever term
alone supplies $Pa^3/(3EI) = 2000\cdot 1.728/(2.4\times 10^{6}) \approx
1.44\,\si{\milli\meter}$; the corner-rotation term adds the remaining
$\sim 2.9\,\si{\milli\meter}$. The vertical member, though it carries
only a constant moment, contributes two-thirds of the tip movement, a
result a cantilever-only estimate would miss by a factor of three.

\textbf{Step 5: cross-check by a direct method.} The same deflection
follows from the moment-area or unit-load route without Castigliano, and
agreement between two independent methods is the discipline's standard
verification. Apply a virtual unit vertical load at the tip; its bending
moments are $m_1(x) = x$ on the horizontal member and $m_2(y) = a$ on the
vertical member, the real moments with $P$ set to $1$. The unit-load
deflection integral is
\[
\delta_\text{tip}
= \int_0^a \frac{M_1 m_1}{EI}\,dx + \int_0^b \frac{M_2 m_2}{EI}\,dy
= \int_0^a \frac{(Px)(x)}{EI}\,dx + \int_0^b \frac{(Pa)(a)}{EI}\,dy
= \frac{P a^3}{3EI} + \frac{P a^2 b}{EI},
\]
identical to the Castigliano result term for term. Two methods, one
number; the analysis is closed. The dropped axial term in the vertical
member would add $PL_2/(EA)$ at the order of $P b/(EA) \approx 2000 \cdot
0.8/(200\times 10^9 \cdot 9\times 10^{-4}) \approx 9\times 10^{-9}\,
\si{\meter}$, six orders of magnitude below the bending deflection, which
justifies dropping it for this slender frame.

\subsection{Worked example: deflection of a truss joint by the unit-load
method}

The unit-load method that cross-checked the frame is the standard tool
for the deflection of a pin-jointed truss, where the members carry only
axial force and the bending integrals collapse to a sum over bars. The
deflection of a chosen joint in a chosen direction is
\[
\delta = \sum_i \frac{n_i N_i L_i}{E_i A_i},
\]
with $N_i$ the real axial force in bar $i$ under the actual load, $n_i$
the axial force in the same bar under a virtual unit load applied at the
joint in the wanted direction, and $L_i/(E_i A_i)$ the bar's axial
flexibility. Consider the two-bar truss of
figure~\ref{fig:vol03ch05:truss-unit-load}: a horizontal bar $BC$ of
length $a = 3.0\,\si{\meter}$ and a diagonal bar $AC$ running from a
higher wall pin $A$ down to the loaded joint $C$, with vertical rise
$h = 2.25\,\si{\meter}$, so the diagonal length is $L_{AC} =
\sqrt{a^2 + h^2} = 3.75\,\si{\meter}$. A vertical load $P =
15\,\si{\kilo\newton}$ hangs at $C$. Both bars are steel with
$EA = 4.0\times 10^{7}\,\si{\newton}$.

\input{volumes/03-force/ch05-energy-methods/figures/fig-truss-unit-load}

\textbf{Real bar forces.} Resolve at joint $C$. The diagonal makes an
angle $\phi$ with the horizontal where $\sin\phi = h/L_{AC} = 2.25/3.75 =
0.6$ and $\cos\phi = a/L_{AC} = 3.0/3.75 = 0.8$. Vertical equilibrium at
$C$ requires the diagonal to carry the load: $N_{AC}\sin\phi = P$, so
$N_{AC} = P/0.6 = 25\,\si{\kilo\newton}$ (tension). Horizontal
equilibrium requires the horizontal bar to balance the diagonal's
horizontal pull: $N_{BC} = -N_{AC}\cos\phi = -25\cdot 0.8 =
-20\,\si{\kilo\newton}$ (compression).

\textbf{Virtual unit-load forces.} A virtual unit downward load at $C$
produces, by the identical statics with $P \to 1$, $n_{AC} = 1/0.6 =
1.667$ and $n_{BC} = -1.333$.

\textbf{Assemble the deflection.}
\[
\delta_C = \frac{n_{AC} N_{AC} L_{AC}}{EA}
         + \frac{n_{BC} N_{BC} L_{BC}}{EA}.
\]
The two products are $n_{AC}N_{AC} = 1.667\cdot 25000 = 41{,}670\,
\si{\newton}$ over length $3.75\,\si{\meter}$, and $n_{BC}N_{BC} =
(-1.333)(-20000) = 26{,}670\,\si{\newton}$ over length
$3.0\,\si{\meter}$. Both products are positive: the compression bar
contributes a positive deflection because both its real and virtual
forces are compressive and their product is positive. So
\[
\delta_C = \frac{41{,}670 \cdot 3.75 + 26{,}670 \cdot 3.0}
                {4.0\times 10^{7}}
= \frac{156{,}300 + 80{,}000}{4.0\times 10^{7}}
\approx 5.9\times 10^{-3}\,\si{\meter},
\]
a downward joint deflection of about $5.9\,\si{\milli\meter}$. The method
delivered one deflection component from two statics passes and a sum over
two bars, with no deformed-geometry sketch and no compatibility equation.

\subsection{End-to-end case study: energy absorption in a drop impact}

Energy methods carry their largest practical weight in impact, where the
peak force a structure must survive is many times the static weight of
the striking mass and where a direct force analysis would need the full
time history of the collision. The energy route sidesteps the time
history: equate the work the falling mass does to the strain energy the
structure stores at peak deflection, and the peak force and deflection
follow from one scalar balance.

\textbf{The setup.} A mass $m = 50\,\si{\kilogram}$ is dropped from a
height $h = 0.30\,\si{\meter}$ onto the free tip of a steel cantilever of
length $L = 1.0\,\si{\meter}$ and bending stiffness $EI = 3.0\times
10^{4}\,\si{\newton\meter\squared}$ (a light beam, chosen so the impact
is severe). Find the peak deflection, the peak force on the beam, and the
dynamic load factor by which the impact exceeds a static placement of the
same mass.

\textbf{Static reference.} Placed gently at the tip, the weight $W = mg =
50 \cdot 9.81 = 491\,\si{\newton}$ produces the static tip deflection of
a cantilever,
\[
\delta_\text{st} = \frac{W L^3}{3 EI}
= \frac{491 \cdot 1.0^3}{3 \cdot 3.0\times 10^{4}}
\approx 5.45\times 10^{-3}\,\si{\meter},
\]
about $5.5\,\si{\milli\meter}$. The tip behaves as a linear spring of
stiffness $k = W/\delta_\text{st} = 3EI/L^3 = 9.0\times 10^{4}\,
\si{\newton\per\meter}$.

\textbf{Energy balance.} The mass falls the drop height $h$ plus the
extra distance $\delta_\text{max}$ it pushes the tip down before
stopping, doing total work $W(h + \delta_\text{max})$. At peak deflection
the mass is momentarily at rest, so all of that work is stored as strain
energy $\tfrac12 k\,\delta_\text{max}^2$ in the beam:
\[
W(h + \delta_\text{max}) = \tfrac12 k\,\delta_\text{max}^2.
\]
Writing the static deflection $\delta_\text{st} = W/k$ and rearranging
gives a quadratic in $\delta_\text{max}$ whose physical root is
\[
\delta_\text{max} = \delta_\text{st}\left(1 + \sqrt{1 +
\frac{2h}{\delta_\text{st}}}\right) = n\,\delta_\text{st},
\]
with the dimensionless \engterm{dynamic load factor}
\[
n = 1 + \sqrt{1 + \frac{2h}{\delta_\text{st}}}.
\]
The factor depends only on the ratio of drop height to static deflection,
plotted in figure~\ref{fig:vol03ch05:impact-factor}.

\input{volumes/03-force/ch05-energy-methods/figures/fig-impact-factor}

\textbf{Numbers.} The drop-height ratio is $2h/\delta_\text{st} =
2\cdot 0.30/5.45\times 10^{-3} = 110$, so
\[
n = 1 + \sqrt{1 + 110} = 1 + \sqrt{111} \approx 1 + 10.5 = 11.5.
\]
The peak deflection is $\delta_\text{max} = 11.5 \cdot 5.45\times 10^{-3}
\approx 6.3\times 10^{-2}\,\si{\meter}$, about $63\,\si{\milli\meter}$,
and the peak force is $F_\text{max} = k\,\delta_\text{max} = n\,W = 11.5
\cdot 491 \approx 5.6\,\si{\kilo\newton}$.

\textbf{Verdict.} A $50\,\si{\kilogram}$ mass that weighs half a
kilonewton when set down gently strikes the beam with more than five and
a half kilonewtons when dropped a third of a metre, an eleven-fold
amplification. A designer who sized the beam for the static weight alone
would under-rate it by a factor of eleven, and the beam would yield or
fail on the first drop. The amplification grows with the square root of
the drop height and is worst for stiff members, where $\delta_\text{st}$
is small and the ratio $2h/\delta_\text{st}$ large. The lesson the energy
method makes unavoidable is that a stiffer structure is not always a
safer one under impact: stiffness lowers the static deflection but raises
the dynamic load factor, and a compliant element deliberately added in
series (a rubber pad, a crush tube, a spring mount) lowers the peak force
by giving the impact energy a larger deflection over which to be
absorbed. This is the design principle behind crumple zones, packaging
foam, and the energy-absorbing mounts on sensitive equipment.

\begin{principle}[Impact amplifies the static response]
A load applied dynamically deflects and stresses an elastic member by a
dynamic load factor $n = 1 + \sqrt{1 + 2h/\delta_\text{st}}$ over the
static value, where $h$ is the drop height and $\delta_\text{st}$ the
static deflection under the same weight. A suddenly applied load ($h=0$)
already doubles the response. Sizing a member against the static load
alone is unsafe whenever the load can arrive with any velocity; the
energy balance between work done and strain energy stored gives the peak
without the collision's time history.
\end{principle}

\subsection{End-to-end case study: Rayleigh-Ritz estimate of a column
buckling load}

The energy method estimates a buckling load by the same quotient that
estimates a natural frequency, and the calculation carried end to end shows
where the method is sharp and where a careless trial shape misleads. A
straight elastic column of length $L$ and bending stiffness $EI$, pinned at
both ends, carries an axial compression $P$. Below a critical value the
straight column is the only equilibrium; at the critical value a buckled
shape becomes possible alongside it, and the column fails by lateral
deflection rather than by crushing. We want that critical load, and we want
it without solving the governing differential equation.

\textbf{The energy quotient.} A column on the verge of buckling trades a
small lateral deflection $w(x)$ for a small drop of the load $P$ as the ends
draw together. The bending strain energy stored in the slight curvature is
$U_\text{bend} = \tfrac12\int_0^L EI\,[w''(x)]^2\,dx$, the same form as the
beam strain energy of the section opener. The load $P$ moves down by the
shortening of the column's projected length, $\Delta = \tfrac12\int_0^L
[w'(x)]^2\,dx$, so the load does work $P\Delta$ on the column. Buckling is
possible at the load where the work the load can do just matches the strain
energy a deflection would cost, $U_\text{bend} = P\Delta$, which rearranges
to the Rayleigh quotient for buckling,
\[
P \;\le\; R[w]
= \frac{\displaystyle\int_0^L EI\,[w''(x)]^2\,dx}
       {\displaystyle\int_0^L [w'(x)]^2\,dx},
\]
an over-estimate of the true critical load for any admissible trial shape
$w$, by the same stationarity argument that made the frequency quotient an
over-estimate: constraining the column to one assumed shape stiffens it, and
a stiffer column buckles at a higher load.

\input{volumes/03-force/ch05-energy-methods/figures/fig-column-buckling-modes}

\textbf{Step 1: a one-term parabola.} The crudest admissible shape that
vanishes at both pins is the parabola $w = s(L-s)$, with $s$ the distance
from one end (figure~\ref{fig:vol03ch05:column-buckling-modes}). Its slope is
$w' = L - 2s$ and its curvature is the constant $w'' = -2$. The two integrals
are immediate,
\[
\int_0^L EI\,(w'')^2\,ds = 4EI\,L,
\qquad
\int_0^L (w')^2\,ds = \int_0^L (L - 2s)^2\,ds = \tfrac13 L^3,
\]
so the estimate is
\[
P_\text{cr} \le \frac{4EI\,L}{\tfrac13 L^3} = \frac{12\,EI}{L^2}.
\]
The exact Euler result for a pinned-pinned column is $P_\text{cr} = \pi^2
EI/L^2 = 9.8696\,EI/L^2$, so the one-term parabola over-estimates by
$12/9.8696 = 1.216$, about $22\,\%$ high. The parabola looks almost
indistinguishable from the true half-sine shape in
figure~\ref{fig:vol03ch05:column-buckling-modes}, yet its buckling estimate
is poor, and the reason is instructive: the quotient weights the curvature
$w''$, and the parabola carries constant curvature where the true mode has
curvature peaking at mid-span and falling to zero at the pins. A shape that
is close in displacement can be far off in second derivative, and buckling
reads the second derivative.

\textbf{Step 2: a shape that respects the moment condition.} The bending
moment in a pinned-pinned column vanishes at each pin, so the true mode has
$w'' = 0$ at both ends, a condition the parabola violates. The static
deflection of a uniformly loaded simply supported beam, $w = sL^3 - 2s^3 L +
s^4$, satisfies $w = 0$ and $w'' = 0$ at both ends and is the natural next
trial. Carrying its integrals through gives
\[
P_\text{cr} \le \frac{168}{17}\,\frac{EI}{L^2} = 9.882\,\frac{EI}{L^2},
\]
within $0.13\,\%$ of the exact $\pi^2 EI/L^2$. One extra condition met, the
zero-moment end state, has cut the error from twenty-two per cent to a
tenth of a per cent. The lesson carries past this problem: a trial shape
should satisfy not only the geometric boundary conditions it must (zero
displacement at a pin, zero displacement and slope at a clamp) but, where it
can, the natural boundary conditions on the higher derivatives, because the
energy quotient is sensitive to exactly those.

\textbf{Step 3: a two-term Ritz expansion and convergence.} The Ritz
refinement of the Rayleigh idea writes the trial shape as a sum of admissible
functions with adjustable weights, $w = a_1\phi_1 + a_2\phi_2$, and makes the
quotient stationary with respect to the weights. With $\phi_1 = s(L-s)$ and
$\phi_2 = s^2(L-s)^2$, stationarity reduces to the generalised eigenvalue
problem $(\mathbf{K} - P\,\mathbf{G})\,\mathbf{a} = 0$ with the
curvature-stiffness matrix $\mathbf{K}_{ij} = \int_0^L \phi_i''\phi_j''\,ds$
and the geometric matrix $\mathbf{G}_{ij} = \int_0^L \phi_i'\phi_j'\,ds$.
Evaluating the integrals,
\[
\mathbf{K} = \begin{pmatrix} 4L & 0 \\ 0 & \tfrac45 L^5 \end{pmatrix},
\qquad
\mathbf{G} = \begin{pmatrix} \tfrac13 L^3 & \tfrac{1}{15} L^5 \\[2pt]
\tfrac{1}{15} L^5 & \tfrac{2}{105} L^7 \end{pmatrix}
\frac{1}{EI},
\]
and the smaller root of the two-by-two determinant condition is
\[
P_\text{cr} \le 2\bigl(45 - \sqrt{1605}\bigr)\,\frac{EI}{L^2}
= 9.8751\,\frac{EI}{L^2},
\]
within $0.056\,\%$ of the exact value, and the larger root estimates the
second buckling mode (the load at which the column would buckle into a full
sine wave if the first mode were restrained). The Ritz expansion converges
to the exact load monotonically from above as more terms are added: each new
admissible function relaxes the artificial stiffening and lowers the
estimate toward the truth.

\textbf{Step 4: a number, and a design check.} Take a steel strut of length
$L = 3.0\,\si{\meter}$ and a thin-walled square section with $I = 2.0\times
10^{-7}\,\si{\meter\tothe{4}}$, so $EI = 200\times 10^{9}\cdot 2.0\times
10^{-7} = 4.0\times 10^{4}\,\si{\newton\meter\squared}$. The exact critical
load is
\[
P_\text{cr} = \frac{\pi^2 EI}{L^2}
= \frac{9.8696\cdot 4.0\times 10^{4}}{3.0^2}
\approx 4.39\times 10^{4}\,\si{\newton},
\]
about $44\,\si{\kilo\newton}$. The two-term Ritz estimate
$9.8751\,EI/L^2 \approx 4.389\times 10^{4}\,\si{\newton}$ lands on top of it;
the one-term parabola would have read $12\,EI/L^2 \approx 5.33\times
10^{4}\,\si{\newton}$, a $22\,\%$ over-estimate that, taken as a capacity,
would size the strut unsafely. The number to carry is the discipline: a
Rayleigh-Ritz buckling estimate is always an upper bound on the capacity, so
an over-estimate is unconservative, and the engineer who uses the method for
a capacity check must either refine the trial shape until it converges or
treat the crude one-term value as a ceiling to stay well under, not as the
allowable load.

\subsection{End-to-end case study: snap-through of a shallow arch}

A second structural instability lives entirely in the shape of a potential
energy curve and shows the energy method finding a critical load that no
linear analysis can see. A shallow two-bar truss, the von Mises truss, has
two pin-ended elastic bars meeting at an apex a small height $h_0$ above the
line joining the supports, with horizontal half-span $b$ so each bar has
length $\ell_0 = \sqrt{b^2 + h_0^2}$. A downward load $P$ at the apex pushes
it down by a distance $w$. For a rise $h_0$ small against the span, the
structure can flip from its initial upward shape to an inverted downward
shape, a snap-through, and the load at which it does is a limit load that a
small-deflection analysis misses entirely because the flip is a large-
displacement event.

\textbf{Step 1: the total potential energy.} As the apex drops by $w$, the
current rise is $h_0 - w$ and each bar shortens (then re-lengthens) by
$\ell_0 - \ell(w)$ with $\ell(w) = \sqrt{b^2 + (h_0 - w)^2}$. For a shallow
truss the bar elongation is well approximated by $e(w) \approx
\bigl[(h_0 - w)^2 - h_0^2\bigr]/(2b) = (w^2 - 2h_0 w)/(2b)$. The axial strain
energy of the two bars, each of stiffness $EA/\ell_0 \approx EA/b$, is
$U_\text{bar} = 2\cdot\tfrac12 (EA/b)\,e(w)^2$, and the load contributes a
potential $-Pw$. The total potential energy is
\[
\Pi(w) = \frac{EA}{b}\,\frac{(w^2 - 2h_0 w)^2}{4b^2} - P\,w
= \frac{EA}{4b^3}\,(w^2 - 2h_0 w)^2 - P\,w.
\]
Figure~\ref{fig:vol03ch05:snap-through} draws the shape of this energy at
three load levels: a symmetric double well with no load, a tilted pair of
wells under a modest load, and, at the critical load, a single inflection
where the shallow well and the barrier between the wells merge.

\input{volumes/03-force/ch05-energy-methods/figures/fig-snap-through}

\textbf{Step 2: the equilibrium path.} Equilibrium is the stationarity of
$\Pi$, $d\Pi/dw = 0$, which gives the load that holds the apex at any
displacement $w$,
\[
P(w) = \frac{EA}{b^3}\bigl(w^3 - 3h_0 w^2 + 2h_0^2 w\bigr).
\]
This is the load-displacement path of
figure~\ref{fig:vol03ch05:snap-load-path}. The path rises, turns over at a
limit point, falls through a stretch of negative slope, and rises again, a
shape no linear stiffness can produce.

\input{volumes/03-force/ch05-energy-methods/figures/fig-snap-load-path}

\textbf{Step 3: the limit load.} The limit points are where the path's slope
vanishes, $dP/dw = 0$, that is where the tangent stiffness drops to zero and
the structure can deflect with no increase in load. Differentiating,
$3w^2 - 6h_0 w + 2h_0^2 = 0$, with roots
\[
w_\text{lim} = h_0\left(1 \mp \frac{1}{\sqrt3}\right).
\]
The upper limit point, at $w_1 = h_0(1 - 1/\sqrt3) \approx 0.423\,h_0$,
carries the snap load
\[
P_\text{max} = \frac{2\sqrt3}{9}\,\frac{EA\,h_0^3}{b^3}
\approx 0.385\,\frac{EA\,h_0^3}{b^3}.
\]
The cubic dependence on the rise $h_0$ is the structural signature of
snap-through: a flatter arch snaps at a far lower load, and a perfectly flat
truss ($h_0 \to 0$) carries no snap load at all because there is no upward
shape to flip out of.

\textbf{Step 4: a number and the snap.} Take bars with $EA = 2.0\times
10^{6}\,\si{\newton}$, a half-span $b = 0.50\,\si{\meter}$, and an initial
rise $h_0 = 50\,\si{\milli\meter}$, a rise-to-span ratio of one in ten. The
snap load is
\[
P_\text{max} = 0.385\cdot\frac{2.0\times 10^{6}\cdot (0.050)^3}{(0.50)^3}
= 0.385\cdot\frac{2.0\times 10^{6}\cdot 1.25\times 10^{-4}}{0.125}
\approx 7.7\times 10^{2}\,\si{\newton},
\]
about $770\,\si{\newton}$, reached at an apex drop $w_1 \approx 0.423\cdot 50
= 21\,\si{\milli\meter}$. Push the apex past that $21\,\si{\milli\meter}$
under load control and the structure cannot hold: the tangent stiffness has
gone negative, the apex accelerates down through the unstable branch, and it
snaps to the inverted shape near $w \approx 2h_0 = 100\,\si{\milli\meter}$,
arriving with the energy released by the drop converted to kinetic energy and
dissipated in the bars and supports. The dashed branch of
figure~\ref{fig:vol03ch05:snap-load-path} is a real continuum of equilibria,
each one a maximum of $\Pi$ in the apex coordinate and so unstable; a load-
controlled test jumps across it, while a displacement-controlled test (an
apex driven down at a prescribed rate) can trace the whole curve, including
the descending branch, because the prescribed displacement supplies the
missing stiffness.

\textbf{Step 5: why the linear analysis is blind.} A small-deflection
stiffness analysis of this truss linearises $P(w)$ about $w = 0$ and reports
the initial tangent stiffness $dP/dw|_0 = 2EA\,h_0^2/b^3$, a positive number
that predicts a stable structure stiffening under load without bound. The
limit load lives in the cubic term that the linearisation discards, so the
linear model never sees the snap. The energy method finds it because the
method works with the full potential energy and reads stability off the
curvature of $\Pi$, which changes sign at the limit point. This is the
recurring value of the energy formulation for stability: a quantity that a
force balance treats as a small correction, the change in geometry under
load, is exactly the quantity that controls whether the structure stands or
snaps. Snap-through is the mechanism behind the click of a bistable switch,
the oil-canning of a thin panel, the action of a jumping popper toy, and the
collapse of shallow domes and reticulated shells under concentrated load,
where it is a recognised and sometimes catastrophic failure mode.

\begin{principle}[Limit loads live in the geometry, not the stiffness]
A structure whose equilibrium depends on its deformed geometry, a shallow
arch, a slender strut, a cable net, can lose stability at a limit load where
its tangent stiffness drops to zero, with no material having yielded and no
member having broken. The limit load is invisible to a small-deflection
analysis that linearises about the undeformed shape, because it lives in the
geometric (large-displacement) terms that the linearisation discards. The
energy method finds it by tracking the total potential energy as a function
of the deformed configuration and locating where its curvature changes sign.
\end{principle}

\begin{mastery}
\textbf{The second variation, stability, and Rayleigh-Ritz as the seed of
the finite-element method.} The principle of stationary potential energy
locates an equilibrium as a stationary point of $\Pi$, but stationarity alone
does not distinguish a stable equilibrium from an unstable one; the wells,
saddles, and peaks of the snap-through landscape are all stationary points.
The discriminating quantity is the second variation. Expand the potential
energy about an equilibrium configuration $q_0$ in a small admissible
perturbation $\eta$,
\[
\Pi(q_0 + \eta) = \Pi(q_0)
+ \underbrace{\delta\Pi}_{=\,0\ \text{at equilibrium}}
+ \tfrac12\,\delta^2\Pi + \cdots,
\]
where the first variation $\delta\Pi$ vanishes because $q_0$ is an
equilibrium and the second variation $\delta^2\Pi$ is a quadratic form in
$\eta$. The equilibrium is stable when $\delta^2\Pi > 0$ for every admissible
$\eta$, that is when the potential energy is a strict local minimum;
unstable when $\delta^2\Pi$ can be made negative by some perturbation,
the configuration sitting at a saddle or a maximum; and marginal, the
critical condition, when $\delta^2\Pi$ first admits a perturbation that makes
it vanish. For a discrete system the second variation is the quadratic form
$\tfrac12\,\eta^{\mathsf T}\mathbf{H}\,\eta$ with $\mathbf{H}$ the Hessian of
$\Pi$, and stability is the positive-definiteness of $\mathbf{H}$; the loss
of stability at a limit or bifurcation load is the moment a Hessian
eigenvalue passes through zero, the eigenvector being the buckling or
snap-through mode shape. The single-coordinate criterion $U'' > 0$ of the
spring example and the matrix criterion are one idea at two sizes.

The same quadratic structure is the bridge from the energy principle to
modern computation. The Rayleigh-Ritz method approximates a continuous
structure's deflection by a finite sum $w = \sum_k a_k\phi_k$ of admissible
shape functions and makes the potential energy stationary with respect to the
weights $a_k$, turning a variational problem over a function space into an
algebraic problem in finitely many unknowns. The stationarity conditions
$\partial\Pi/\partial a_k = 0$ are a linear system $\mathbf{K}\mathbf{a} =
\mathbf{f}$ for a static problem, or a generalised eigenvalue problem
$(\mathbf{K} - \lambda\mathbf{G})\mathbf{a} = 0$ for a buckling or vibration
problem, exactly the two-by-two systems the column-buckling and
double-pendulum analyses produced by hand. The finite-element method is
Rayleigh-Ritz with one decisive change in the choice of shape functions:
instead of global functions spanning the whole structure, it uses shape
functions that are non-zero only over one small element of a mesh, so the
stiffness matrix $\mathbf{K}$ becomes sparse and the method scales to
millions of unknowns. Every finite-element structural solver in service is a
Rayleigh-Ritz minimisation of a discretised potential energy, and the
positive-definiteness of its stiffness matrix is the discrete second-variation
stability condition. The hand calculations of this chapter are the same
machine in miniature.
\end{mastery}

\subsection{Worked example: Rayleigh estimate of a natural frequency}

The Rayleigh method of the section 5.5 mastery box estimates a vibrating
system's fundamental frequency from an energy ratio on an assumed mode
shape, the practising engineer's fastest check on a
structural-dynamics result. Estimate the
fundamental frequency of a uniform cantilever beam of length $L$, mass
per unit length $\rho A$, and bending stiffness $EI$, carrying no tip
mass.

\textbf{Trial shape.} Choose a shape that satisfies the geometric
(clamped-end) boundary conditions $\phi(0) = 0$ and $\phi'(0) = 0$. The
static-deflection shape under a uniform load is one good choice; a
simpler admissible polynomial is $\phi(x) = 1 - \cos(\pi x / (2L))$,
which gives $\phi(0) = 0$, $\phi'(0) = 0$, and a tip value $\phi(L) = 1$.

\textbf{Energy ratio.} The Rayleigh quotient for a beam is the ratio of
peak bending strain energy to a reference kinetic energy,
\[
\omega^2 = \frac{\displaystyle\int_0^L EI\,[\phi''(x)]^2\,dx}
                {\displaystyle\int_0^L \rho A\,[\phi(x)]^2\,dx}.
\]
With $\phi = 1 - \cos(\pi x/(2L))$, the second derivative is $\phi'' =
(\pi/(2L))^2\cos(\pi x/(2L))$. The numerator integral evaluates to
$EI\,(\pi/(2L))^4 \cdot (L/2)$ and the denominator to $\rho A\,L\,(3/2 -
4/\pi)$, giving
\[
\omega_1 \approx 3.66\,\sqrt{\frac{EI}{\rho A L^4}}.
\]
The exact Euler-Bernoulli coefficient is $(1.875)^2 = 3.516$, so the
crude one-term trial shape over-estimates the fundamental by about
$4\,\%$, on the right side: the Rayleigh estimate is always an
upper bound, because constraining the motion to a single assumed shape
stiffens the system. A designer who reads $4.6\,\si{\hertz}$ from a
finite-element model and computes a Rayleigh estimate of $4.8\,\si{\hertz}$
has a consistent pair; a model returning $1\,\si{\hertz}$ or
$15\,\si{\hertz}$ against this estimate carries an error in the model, not
a surprise in the structure.

\subsection{Worked example: equilibrium of a spring system by minimum
potential energy}

The static counterpart of Hamilton's principle is the principle of
\engterm{minimum potential energy}: a conservative system in stable
equilibrium sits at a minimum of its total potential energy, and the
equilibrium configuration is found by setting the energy's gradient to
zero. The method recovers force balances from a single scalar without
drawing free-body diagrams, and it extends directly to systems where a
force balance would be awkward.

A mass $m$ hangs from a spring of stiffness $k$ fixed to a support
(figure~\ref{fig:vol03ch05:min-pe-springs}). Let $x$ be the downward
displacement of the mass from the spring's natural length. The total
potential energy is the spring energy plus the gravitational energy,
\[
U(x) = \tfrac12 k x^2 - m g x,
\]
the gravitational term negative because the mass loses height as $x$
grows. The equilibrium is the stationary point,
\[
\frac{dU}{dx} = k x - m g = 0
\quad\Longrightarrow\quad
x_\text{eq} = \frac{mg}{k},
\]
the familiar static stretch, recovered without a force balance. The
second derivative $U'' = k > 0$ confirms the stationary point is a
minimum, so the equilibrium is stable; the curvature $k$ is the system's
stiffness and sets the small-oscillation frequency $\omega = \sqrt{k/m}$
about the equilibrium. The same procedure handles a mass held between two
springs of stiffnesses $k_1$ and $k_2$: the energy is $\tfrac12 k_1 x^2 +
\tfrac12 k_2 x^2 - F x$ under an applied force $F$, the gradient gives
$x_\text{eq} = F/(k_1 + k_2)$ (springs in parallel add their
stiffnesses), and the curvature $k_1 + k_2$ confirms stability. The
energy method scales to many springs and many masses with no new idea:
write the total potential energy, set its gradient to zero for the
equilibrium, and read stability off the curvature matrix.

\input{volumes/03-force/ch05-energy-methods/figures/fig-min-pe-springs}

\subsection{Worked example: work-energy accounting with friction}

Friction breaks the conservation of mechanical energy, and the
energy method handles it through a single change to the balance: the
mechanical energy at the end equals the mechanical energy at the start
minus the work dissipated by friction. A block of mass $m =
4.0\,\si{\kilogram}$ is launched up a rough incline of angle
$\alpha = 25^{\circ}$ with initial speed $v_0 = 6.0\,\si{\meter\per
\second}$; the coefficient of kinetic friction is $\mu = 0.30$. How far
up the incline does the block travel before stopping, and how fast is it
moving when it returns to the launch point?

\textbf{Going up.} The block rises a distance $d$ along the incline,
gaining height $d\sin\alpha$ and losing all its kinetic energy to gravity
and friction. The friction force is $\mu m g\cos\alpha$, opposing the
upward motion, so the energy balance from launch to the highest point is
\[
\tfrac12 m v_0^2 = m g d \sin\alpha + \mu m g \cos\alpha\, d.
\]
The mass cancels, and
\[
d = \frac{v_0^2}{2 g(\sin\alpha + \mu\cos\alpha)}
= \frac{6.0^2}{2 \cdot 9.81 (\,0.423 + 0.30\cdot 0.906\,)}
= \frac{36}{2\cdot 9.81 \cdot 0.695}
\approx 2.64\,\si{\meter}.
\]

\textbf{Coming down.} On the descent gravity drives the block down and
friction now opposes the downward motion, so the energy released by the
descent is the gravitational drop minus the friction loss over the same
distance $d$:
\[
\tfrac12 m v_f^2 = m g d \sin\alpha - \mu m g \cos\alpha\, d.
\]
Solving for the return speed,
\[
v_f = \sqrt{2 g d(\sin\alpha - \mu\cos\alpha)}
= \sqrt{2 \cdot 9.81 \cdot 2.64 (\,0.423 - 0.272\,)}
\approx 2.80\,\si{\meter\per\second}.
\]
The block returns at $2.8\,\si{\meter\per\second}$, well under its launch
speed of $6.0\,\si{\meter\per\second}$, because friction took its toll
twice, once on the way up and once on the way down. The total mechanical
energy lost to friction over the round trip is the kinetic-energy deficit
$\tfrac12 m(v_0^2 - v_f^2) = \tfrac12\cdot 4.0\,(36 - 7.84) \approx
56\,\si{\joule}$, which equals $2\mu m g\cos\alpha\,d = 2\cdot 0.30\cdot
4.0\cdot 9.81\cdot 0.906\cdot 2.64 \approx 56\,\si{\joule}$, the friction
work over the full $2d$ of travel. The two figures agree, which closes
the accounting: every joule the block started with is either returned as
kinetic energy or charged to friction.

\subsection{Worked example: Castigliano deflection of a curved member}

The strain-energy machinery handles a curved member as readily as a straight
one, with the energy integral running over arc length and the bending moment
written as a function of position around the curve. A semicircular member of
radius $R$ and uniform bending stiffness $EI$ is built in at its top end and
free at its bottom, carrying a vertical load $P$ at the free end
(figure~\ref{fig:vol03ch05:curved-member}). We want the vertical deflection
of the loaded end.

\input{volumes/03-force/ch05-energy-methods/figures/fig-curved-member}

\textbf{Bending moment around the arc.} Locate a generic cross-section by the
angle $\phi$ measured from the fixed end, so the section sits at horizontal
distance $R\sin\phi$ from the vertical line through the load. The load $P$ is
vertical, so its moment arm about the section is that horizontal distance,
and the bending moment is
\[
M(\phi) = P\,R\sin\phi,\qquad 0 \le \phi \le \pi.
\]
The moment is zero at the fixed end ($\phi = 0$) and at the bottom of the arc
($\phi = \pi$), and peaks at the side of the arc ($\phi = \pi/2$) where the
horizontal lever arm is the full radius. Axial and shear contributions to a
slender curved member are small and dropped, as for the straight frame.

\textbf{Strain energy and Castigliano.} The bending strain energy integrates
the moment around the arc, with the arc-length element $ds = R\,d\phi$,
\[
U = \int_0^\pi \frac{M(\phi)^2}{2EI}\,R\,d\phi
  = \frac{R}{2EI}\int_0^\pi P^2 R^2 \sin^2\phi\,d\phi.
\]
The vertical tip deflection is Castigliano's derivative with respect to $P$,
which differentiates inside the integral,
\[
\delta = \frac{\partial U}{\partial P}
       = \frac{R}{EI}\int_0^\pi P R^2 \sin^2\phi\,d\phi
       = \frac{P R^3}{EI}\int_0^\pi \sin^2\phi\,d\phi
       = \frac{P R^3}{EI}\cdot\frac{\pi}{2}
       = \frac{\pi P R^3}{2EI},
\]
using $\int_0^\pi \sin^2\phi\,d\phi = \pi/2$. The curved member is more
flexible than a straight cantilever of the same material because its bending
moment is distributed around a curve rather than concentrated, and the
$R^3$ scaling matches the straight-cantilever $L^3$ scaling with the radius
playing the role of the length.

\textbf{A number.} Take $R = 0.40\,\si{\meter}$, a load $P =
1.0\,\si{\kilo\newton}$, and $EI = 8.0\times 10^{4}\,\si{\newton\meter
\squared}$ (a small steel tube). Then
\[
\delta = \frac{\pi\cdot 1000\cdot 0.40^3}{2\cdot 8.0\times 10^{4}}
= \frac{\pi\cdot 1000\cdot 0.064}{1.6\times 10^{5}}
\approx 1.26\times 10^{-3}\,\si{\meter},
\]
about $1.3\,\si{\milli\meter}$. The horizontal deflection of the same tip
follows from the same machinery with a fictitious horizontal load $Q$ added
at the free end, the horizontal moment arm written into $M$, and the
derivative $\partial U/\partial Q$ taken at $Q = 0$, the standard
dummy-load trick for a deflection in a direction with no real applied force.

\subsection{Worked example: an indeterminate beam by least work}

Castigliano's machinery resolves a statically indeterminate structure
through the principle of least work: the redundant reactions of a linearly
elastic structure take the values that make the strain energy stationary
with respect to them. A propped cantilever, built into a wall at one end and
resting on a simple prop at the other, spanning $L$ under a uniform load $w$
per unit length, has one redundant reaction; statics alone gives three
equilibrium equations and four unknown reactions, so the structure is
indeterminate to the first degree, and the energy method supplies the
missing equation.

\textbf{Choose the redundant.} Take the upward prop reaction $R_B$ at the
simply supported end as the redundant. Measuring $x$ from the prop toward the
wall, the bending moment at a section combines the prop reaction acting up
through the lever arm $x$ and the distributed load acting down,
\[
M(x) = R_B\,x - \tfrac12 w x^2,\qquad 0 \le x \le L,
\]
a moment written entirely in terms of $R_B$ and the known load, with the
wall reactions never introduced.

\textbf{Least work supplies the compatibility condition.} The prop is rigid,
so the beam's deflection there is zero, and Castigliano's theorem reads that
deflection as $\partial U/\partial R_B$. Setting it to zero is the principle
of least work,
\[
\frac{\partial U}{\partial R_B}
= \int_0^L \frac{M}{EI}\,\frac{\partial M}{\partial R_B}\,dx
= \frac{1}{EI}\int_0^L \bigl(R_B x - \tfrac12 w x^2\bigr)\,x\,dx = 0.
\]
Carrying out the integral,
\[
\int_0^L \bigl(R_B x^2 - \tfrac12 w x^3\bigr)\,dx
= R_B\,\frac{L^3}{3} - \frac{w L^4}{8} = 0
\quad\Longrightarrow\quad
R_B = \frac{3 w L}{8}.
\]
The prop carries three-eighths of the total load $wL$, and the wall carries
the remaining five-eighths together with a fixing moment $M_A = wL^2/8$ that
follows from statics once $R_B$ is known. The energy method delivered the
redundant from one scalar integral and one stationarity condition, in place
of the compatibility equation a deflection-based analysis would have to set
up by integrating the beam's curvature twice and matching boundary
conditions.

\subsection{Worked example: impact of a moving mass on a spring buffer}

The kinetic-to-strain-energy balance that governed the drop impact governs a
horizontal collision into a spring buffer, the model for a railway-car
coupler, a dock bumper, or a machine end-stop. A car of mass $m = 20\,
\si{\tonne} = 2.0\times 10^{4}\,\si{\kilogram}$ rolls at $v =
1.5\,\si{\meter\per\second}$ into a spring buffer of stiffness $k = 2.5\times
10^{6}\,\si{\newton\per\meter}$. Find the peak buffer compression and the
peak force on the car.

\textbf{Energy balance.} The car's kinetic energy at contact converts
entirely to strain energy stored in the buffer at peak compression, where
the car is momentarily at rest (gravity does no work on the horizontal
motion, and the buffer is taken as undamped over the brief compression):
\[
\tfrac12 m v^2 = \tfrac12 k\,\delta_\text{max}^2
\quad\Longrightarrow\quad
\delta_\text{max} = v\sqrt{\frac{m}{k}}.
\]
The peak force is the spring force at that compression, $F_\text{max} =
k\,\delta_\text{max} = v\sqrt{mk}$, a result that depends on the geometric
mean of mass and stiffness and grows in direct proportion to the closing
speed.

\textbf{Numbers.} The peak compression is
\[
\delta_\text{max} = 1.5\sqrt{\frac{2.0\times 10^{4}}{2.5\times 10^{6}}}
= 1.5\sqrt{8.0\times 10^{-3}}
\approx 0.134\,\si{\meter},
\]
about $134\,\si{\milli\meter}$, and the peak force is
\[
F_\text{max} = 1.5\sqrt{2.0\times 10^{4}\cdot 2.5\times 10^{6}}
= 1.5\sqrt{5.0\times 10^{10}}
\approx 3.35\times 10^{5}\,\si{\newton},
\]
about $335\,\si{\kilo\newton}$, roughly $1.7$ times the car's static weight.
The design trade is the mirror of the drop-impact lesson: a softer buffer
(smaller $k$) compresses further but peaks at a lower force, since
$\delta_\text{max} \propto k^{-1/2}$ while $F_\text{max}\propto k^{+1/2}$, so
halving the stiffness raises the travel by $41\,\%$ and cuts the peak force
by $29\,\%$. A buffer is specified by the largest force the car structure can
take and the longest travel the installation allows, and the two together
fix the stiffness through this energy balance.

\subsection{End-to-end case study: a crumple zone sized by energy
absorption}

The impact balances of the drop and the buffer treat the absorbing element
as a linear spring, where the force rises in proportion to the deflection
and the energy lands as $\tfrac12 k\delta^2$. A vehicle front structure is
designed to do the opposite: it crushes at a force as nearly constant as
the folding of thin-walled steel allows, so that the deceleration it
imposes on the occupant cell stays flat across the whole crush stroke
rather than spiking. The energy method sizes the structure from one
balance, the kinetic energy to be removed against the work the crush force
does over its stroke, and the design lives or dies on a single ratio: the
mean crush force and the available stroke must remove all the energy before
the occupant cell runs out of room.

\textbf{The requirement.} A vehicle of mass $m = 1500\,\si{\kilogram}$
strikes a rigid barrier at $v_0 = 15.6\,\si{\meter\per\second}$
(about $56\,\si{\kilo\meter\per\hour}$, the speed of the standard frontal
crash test). The kinetic energy to be absorbed is
\[
E_k = \tfrac12 m v_0^2 = \tfrac12 \cdot 1500 \cdot 15.6^2
\approx 1.83\times 10^{5}\,\si{\joule},
\]
about $183\,\si{\kilo\joule}$. The occupant cell must not decelerate faster
than the body can survive; a sustained deceleration of $20\,g$ over the
event is a common cell-level design limit, with the restraint system
keeping the occupant's own deceleration below that. We size the front
structure's mean crush force and the crush stroke it needs, then check the
occupant's ride-down.

\textbf{Step 1: the crush stroke for a constant-force structure.} An ideal
crush structure exerts a constant mean force $F_\text{mean}$ over a crush
stroke $s_c$, doing work $F_\text{mean}\,s_c$ on the decelerating mass.
Equating that work to the kinetic energy, and writing the mean deceleration
as $a = F_\text{mean}/m$, the constant-force kinematics give the stroke
directly,
\[
F_\text{mean}\,s_c = \tfrac12 m v_0^2
\quad\Longrightarrow\quad
s_c = \frac{v_0^2}{2\,a},
\]
the same $v^2 = 2 a s$ relation that governs constant-acceleration motion,
read here as a stroke requirement. The stroke depends only on the closing
speed and the deceleration the cell allows, not on the vehicle mass, which
cancels. For the cell limit $a = 20\,g = 196\,\si{\meter\per\second
\squared}$,
\[
s_c = \frac{15.6^2}{2 \cdot 196} \approx 0.62\,\si{\meter},
\]
about $620\,\si{\milli\meter}$ of crush. A typical car carries $0.6$ to
$0.8\,\si{\meter}$ between the front bumper and the firewall, so the
requirement is met but not lavishly; a stiffer cell limit of $15\,g$ would
demand $0.83\,\si{\meter}$, more than most cars have, which is why the
$20\,g$ cell with a softer occupant ride-down is the working split.

\textbf{Step 2: the mean crush force and the structural cross-section.} The
mean force the structure must sustain is
\[
F_\text{mean} = m\,a = 1500 \cdot 196 \approx 2.94\times 10^{5}\,
\si{\newton},
\]
about $294\,\si{\kilo\newton}$, carried by the two front rails together.
Each rail must therefore fold at a mean crush load of roughly
$147\,\si{\kilo\newton}$. A thin-walled square steel rail crushes in an
accordion of folds at a mean force that the established axial-crush relation
ties to the wall thickness $t$, the section side $b$, and the flow stress
$\sigma_0$ of the steel, $F_\text{mean} \approx 13.06\,\sigma_0\,t^{5/3}\,
b^{1/3}$ for a square box. Taking a mild steel flow stress $\sigma_0 \approx
300\,\si{\mega\pascal}$ and a rail side $b = 80\,\si{\milli\meter}$, the
wall thickness that delivers $147\,\si{\kilo\newton}$ per rail is the
solution of
\[
1.47\times 10^{5} = 13.06 \cdot 3.0\times 10^{8}\cdot t^{5/3}\cdot
(0.080)^{1/3},
\]
which gives $t^{5/3} \approx 8.6\times 10^{-5}$ and $t \approx 2.8\times
10^{-3}\,\si{\meter}$, a wall a little under $3\,\si{\milli\meter}$, square
in the range real rails use. The energy balance has sized a sheet-metal
gauge from a crash speed.

\input{volumes/03-force/ch05-energy-methods/figures/fig-crumple-force-deflection}

\input{volumes/03-force/ch05-energy-methods/figures/fig-crumple-survival-space}

\textbf{Step 3: the real force-deflection curve and the stroke efficiency.}
A real rail does not crush at a constant force from the first instant.
Figure~\ref{fig:vol03ch05:crumple-force-deflection} draws the
characteristic: an initial peak as the first section buckles, a trough as
that fold collapses, then a rippling plateau as successive folds form. Two
corrections follow from the shape. The initial peak, two to three times the
mean, must be suppressed, or the early deceleration spike injures the
occupant before the structure settles into folding; crush initiators
(trigger dents, holes, or beads stamped into the rail) lower the first peak
toward the plateau. The usable stroke is shorter than the rail length,
because the folded steel stacks up solid at the end; a crush efficiency of
$\eta \approx 0.7$ (the fraction of the original length that is usable
stroke) is typical, so a rail must be $s_c/\eta \approx 0.62/0.7 \approx
0.89\,\si{\meter}$ long to deliver $0.62\,\si{\meter}$ of effective crush.
The mean force in step 2 is properly read as the area under the real curve
divided by the effective stroke, the equal-area mean drawn in the figure.

\textbf{Step 4: the occupant's share and the survival space.} The front
structure protects the cell; the restraint system protects the occupant
inside it. Figure~\ref{fig:vol03ch05:crumple-survival-space} shows the
split. The cell decelerates at $20\,g$ over the crush stroke $s_c$; an
unrestrained occupant would continue at $v_0$ until striking the
interior, taking the full $v_0$ over the few centimetres of interior
clearance, a lethal deceleration. The belt and airbag couple the occupant
to the decelerating cell and let the occupant ride down through an extra
travel $s_r$ relative to the cell, over which the restraint does work at a
force the body survives. The occupant's own kinetic energy per unit mass is
$\tfrac12 v_0^2$, and the ride-down travel needed to remove it at an
occupant deceleration $a_\text{occ}$ is
\[
s_r = \frac{v_0^2}{2\,a_\text{occ}}.
\]
A restraint that holds the occupant to $40\,g$ (a survivable chest
loading for a belted adult) needs $s_r = 15.6^2/(2\cdot 40\cdot 9.81)
\approx 0.31\,\si{\meter}$ of ride-down. The occupant gets that travel from
the cell's own crush stroke (the cell moves forward into the barrier while
the belt pays out) plus belt stretch and airbag stroke; the budget closes
only if the cell stays rigid, since any intrusion into the survival space is
ride-down travel taken away from the occupant and deceleration handed back.
The case study's recurring lesson is that the crush zone and the occupant
restraint are one energy system: the structure absorbs the vehicle's
energy at the cell limit, the restraint absorbs the occupant's energy at
the body limit, and the survival space is the stroke the two share.

\textbf{Step 5: the energy ledger.} Closing the accounting confirms the
design. The front structure absorbs $F_\text{mean}\,s_c \approx 2.94\times
10^{5}\cdot 0.62 \approx 1.82\times 10^{5}\,\si{\joule}$, almost all
of the $183\,\si{\kilo\joule}$ of vehicle kinetic energy, the small
remainder going to the deformation of the barrier-contact region and to the
parts of the car aft of the rails. The restraint absorbs the occupant's
share, $\tfrac12 m_\text{occ} v_0^2 \approx \tfrac12\cdot 80\cdot 15.6^2
\approx 9.7\times 10^{3}\,\si{\joule}$ for an $80\,\si{\kilogram}$ occupant,
which is already counted inside the vehicle total but must be delivered to
the body at the lower occupant force, not the cell force. Every joule the
car carried into the barrier is charged either to the folding steel or to
the restraint webbing, and the design is the partition of that energy into
strokes and forces each component can take. The same balance sizes a
helmet liner, a packaging foam, a fall-arrest lanyard, and a spacecraft
landing crushable: kinetic energy in, a force the protected thing can
survive, and the stroke that the quotient of the two demands.

\begin{principle}[Constant-force crush minimises the stroke for a force
limit]
A structure that absorbs a fixed kinetic energy $E_k$ at a force not to
exceed $F_\text{max}$ needs the least stroke when it crushes at the
constant force $F_\text{max}$ throughout, giving $s = E_k/F_\text{max}$. Any
force that rises and falls about that mean either exceeds the limit at its
peak or wastes stroke at its trough. The crashworthiness design target is a
flat force plateau over the whole available stroke, achieved by crush
initiators and progressive folding; the same target governs every
energy-absorbing element specified by a survivable force and a finite
travel.
\end{principle}

\subsection{Worked example: strain-energy partition in a stepped torsion
shaft}

A shaft that steps from one diameter to another carries the same internal
torque through both segments but stores its torsional strain energy
unequally, and the partition shows where the compliance of a drivetrain
hides. The shaft of figure~\ref{fig:vol03ch05:stepped-shaft} is fixed at
one end and twisted by a torque $T$ at the other, stepping from a segment of
polar second moment $J_1$ and length $L_1$ to a thinner segment $J_2$,
$L_2$, both of shear modulus $G$. We want the energy stored in each segment
and the total angle of twist at the loaded end.

\input{volumes/03-force/ch05-energy-methods/figures/fig-stepped-shaft}

\textbf{Internal torque and strain energy.} Static equilibrium carries the
applied torque $T$ unchanged through both segments, since no torque is
applied between the ends. The torsional strain energy of a uniform shaft
segment carrying a constant torque is $U = T^2 L/(2GJ)$, the torsional
analogue of the axial $N^2L/(2EA)$, so the two segments store
\[
U_1 = \frac{T^2 L_1}{2 G J_1},
\qquad
U_2 = \frac{T^2 L_2}{2 G J_2},
\qquad
\frac{U_1}{U_2} = \frac{L_1/J_1}{L_2/J_2}.
\]
The energy splits in the ratio of the segments' torsional flexibilities
$L_i/(GJ_i)$. The thin segment, with the smaller $J$, stores the larger
share; a single under-sized step dominates the energy and the compliance of
an otherwise stiff shaft.

\textbf{Angle of twist by Castigliano.} The rotation at the loaded end,
conjugate to the torque $T$, is Castigliano's derivative of the total strain
energy with respect to $T$,
\[
\varphi = \frac{\partial (U_1 + U_2)}{\partial T}
= \frac{T L_1}{G J_1} + \frac{T L_2}{G J_2}
= T\left(\frac{L_1}{G J_1} + \frac{L_2}{G J_2}\right),
\]
the sum of the two segments' twist contributions, exactly the
series-spring addition of torsional flexibilities the figure's caption
names. The energy method recovers the flexibility of a stepped shaft as the
sum of its segments' flexibilities without integrating a twist-rate
equation along the length.

\textbf{A number.} Take a steel shaft with $G = 80\,\si{\giga\pascal}$, a
fat segment of diameter $50\,\si{\milli\meter}$ ($J_1 = \pi d^4/32 =
6.14\times 10^{-7}\,\si{\meter\tothe{4}}$) and length $L_1 = 0.40\,\si
{\meter}$, a thin segment of diameter $30\,\si{\milli\meter}$ ($J_2 =
7.95\times 10^{-8}\,\si{\meter\tothe{4}}$) and length $L_2 = 0.30\,\si
{\meter}$, under a torque $T = 500\,\si{\newton\meter}$. The two
flexibilities are $L_1/(GJ_1) = 0.40/(80\times 10^{9}\cdot 6.14\times
10^{-7}) = 8.1\times 10^{-6}\,\si{\per\newton\per\meter}$ and $L_2/(GJ_2) =
0.30/(80\times 10^{9}\cdot 7.95\times 10^{-8}) = 4.72\times 10^{-5}\,\si
{\per\newton\per\meter}$. The thin segment is nearly six times the more
flexible, so it stores about $85\,\%$ of the energy though it is the shorter
of the two. The total twist is $\varphi = 500\cdot(8.1\times 10^{-6} +
4.72\times 10^{-5}) \approx 2.8\times 10^{-2}\,\si{\radian}$, about
$1.6^\circ$. A designer chasing the compliance would fatten the thin
segment first, where the energy is, not the long fat one where it is not.

\subsection{Applied vignette: a coil spring rate from torsion strain
energy}

The torsion energy of a straight wire delivers the rate of a helical
spring in one balance, a derivation that the catalogue formula
$k = Gd^4/(8D^3 n)$ hides behind a constant. A close-coiled helical
compression spring of mean coil radius $R$ (mean coil diameter $D = 2R$),
wire diameter $d$, and $n$ active coils carries an axial load $F$. The
whole compliance comes from the wire twisting; that the wire happens to be
wound into a helix changes only the bookkeeping of its length, drawn in
figure~\ref{fig:vol03ch05:coil-spring}.

\input{volumes/03-force/ch05-energy-methods/figures/fig-coil-spring}

\textbf{The torque in the wire.} Cut the wire at any section and balance
the part beyond the cut. The axial load $F$ acts along the spring axis at
the perpendicular distance $R$ from the wire centreline, so the section
carries a torque $T = FR$, constant around the whole coil for a
close-coiled spring whose helix angle is small. A direct shear $F$ also
acts on the section, but for a slender wire its energy is smaller than the
torsion energy by a factor of order $(d/D)^2$ and we drop it.

\textbf{The wire length and the strain energy.} The active wire is $n$
coils each of circumference $\pi D$, so its length is $\ell = \pi D n =
2\pi R n$. The polar second moment of the round wire is $J = \pi d^4/32$.
The torsional strain energy stored in the whole wire under the constant
torque $T = FR$ is
\[
U = \frac{T^2\,\ell}{2GJ}
= \frac{(FR)^2\,(2\pi R n)}{2G\,(\pi d^4/32)}
= \frac{32\,F^2 R^3 n}{G\,d^4}.
\]

\textbf{The deflection and the rate.} Castigliano's theorem gives the
axial deflection conjugate to $F$ directly,
\[
\delta = \frac{\partial U}{\partial F}
= \frac{64\,F R^3 n}{G\,d^4}
= \frac{8\,F D^3 n}{G\,d^4},
\]
using $R = D/2$ so $R^3 = D^3/8$. The spring rate is the load per unit
deflection,
\[
k = \frac{F}{\delta} = \frac{G\,d^4}{8\,D^3 n},
\]
the standard catalogue formula, now derived rather than quoted. The
fourth-power dependence on the wire diameter is the dominant design lever:
a wire ten per cent thicker stiffens the spring by forty-six per cent,
while the inverse-cube dependence on the coil diameter and the inverse
dependence on the coil count make a longer, wider spring softer.

\textbf{A number.} Take a steel spring with $G = 80\,\si{\giga\pascal}$,
wire diameter $d = 4.0\,\si{\milli\meter}$, mean coil diameter $D =
30\,\si{\milli\meter}$, and $n = 8$ active coils. The rate is
\[
k = \frac{80\times 10^{9}\cdot (4.0\times 10^{-3})^4}
         {8\cdot (30\times 10^{-3})^3\cdot 8}
= \frac{80\times 10^{9}\cdot 2.56\times 10^{-10}}
       {8\cdot 2.7\times 10^{-5}\cdot 8}
\approx 1.19\times 10^{4}\,\si{\newton\per\meter},
\]
about $11.9\,\si{\kilo\newton\per\meter}$, so a $100\,\si{\newton}$ load
compresses the spring by $\delta = F/k \approx 8.4\,\si{\milli\meter}$ and
stores $U = \tfrac12 F\delta \approx 0.42\,\si{\joule}$. The energy method
turned a one-line torsion result into a spring rate, a deflection, and a
stored energy, with the helix entering only through the wire length.

\subsection{Worked example: a sideways deflection by a fictitious dummy
load}

A deflection in a direction that carries no real applied load is recovered
by the dummy-load trick the derivation named: apply a fictitious load
there, carry it through the moments, differentiate, and set it to zero.
Reuse the L-shaped cantilever frame of the earlier case study, a horizontal
member of length $a$ joined at a rigid corner to a vertical member of length
$b$ built into the wall, with the real vertical tip load $P$. We want the
\emph{horizontal} movement of the tip, for which there is no real
horizontal force.

\textbf{Add the dummy load.} Apply a fictitious horizontal load $Q$ at the
tip, pointing away from the wall, and write the bending moments with both
$P$ and $Q$ present. On the horizontal member, measuring $x$ from the tip,
the vertical load contributes $Px$ as before and the horizontal load $Q$
contributes nothing to bending about the member's own axis (its line of
action runs along the member), so $M_1(x) = Px$. On the vertical member,
measuring $y$ from the corner upward, the vertical load gives the constant
$Pa$ and the horizontal load $Q$ now acts at the growing vertical lever arm
$y$, so
\[
M_2(y) = Pa + Q\,y,\qquad 0\le y\le b.
\]

\textbf{Differentiate and set the dummy load to zero.} The horizontal tip
deflection is Castigliano's derivative with respect to $Q$, evaluated at
$Q=0$,
\[
\delta_\text{horiz}
= \left.\frac{\partial U}{\partial Q}\right|_{Q=0}
= \int_0^a \frac{M_1}{EI}\frac{\partial M_1}{\partial Q}\,dx
+ \int_0^b \frac{M_2}{EI}\frac{\partial M_2}{\partial Q}\,dy.
\]
On the horizontal member $\partial M_1/\partial Q = 0$, so that integral
drops. On the vertical member $\partial M_2/\partial Q = y$, and setting
$Q=0$ leaves $M_2 = Pa$, so
\[
\delta_\text{horiz}
= \int_0^b \frac{(Pa)(y)}{EI}\,dy
= \frac{Pa}{EI}\cdot\frac{b^2}{2}
= \frac{P a b^2}{2EI}.
\]
The horizontal movement comes entirely from the vertical member bending
under the constant moment $Pa$, which rotates and bows it sideways; the
horizontal member contributes nothing because the vertical load puts no
$Q$-sensitive moment in it.

\textbf{A number.} With the earlier frame, $a = 1.2\,\si{\meter}$, $b =
0.8\,\si{\meter}$, $P = 2.0\,\si{\kilo\newton}$, and $EI = 8.0\times 10^{5}
\,\si{\newton\meter\squared}$,
\[
\delta_\text{horiz} = \frac{2000\cdot 1.2\cdot 0.8^2}{2\cdot 8.0\times 10^{5}}
\approx 9.6\times 10^{-4}\,\si{\meter},
\]
about $0.96\,\si{\milli\meter}$ of sideways drift, a fifth of the
$4.3\,\si{\milli\meter}$ vertical drop computed before. The tip of a loaded
frame moves along a slanted path, not straight down, and the dummy load is
how the energy method extracts the component that no real force points
toward.

\subsection{Worked example: a bolt as a spring, and the preload energy}

A tightened bolt clamping a joint is two springs in parallel sharing a
preload, and the energy method fixes how an external load divides between
them, the calculation that decides whether a bolted joint survives fatigue.
The bolt of effective length $L_b$, shank area $A_b$, and modulus $E$ acts
as a tension spring of stiffness $k_b = E A_b/L_b$; the clamped members,
loaded in compression over an effective area $A_m$ and grip length $L_m$,
act as a stiffer compression spring of stiffness $k_m = E_m A_m/L_m$. Both
carry the preload $F_i$ that tightening installs.

\textbf{Stored energy at preload.} Tightening stretches the bolt and
squeezes the members, storing
\[
U_i = \frac{F_i^2}{2 k_b} + \frac{F_i^2}{2 k_m}
= \tfrac12 F_i^2\!\left(\frac{1}{k_b}+\frac{1}{k_m}\right),
\]
the sum of the two springs' energies at the common preload, the bolt
holding the larger share because it is the softer spring. The wrench does
this work against the thread and the joint, and it is recoverable elastic
energy, not loss, until something yields.

\textbf{How an external load divides.} Apply a tensile service load $P$ that
tries to separate the joint. Compatibility requires the bolt and the members
to move together by the same increment $\delta$ as long as the joint stays
closed, so the load splits in proportion to stiffness. The bolt picks up
the share
\[
\Delta F_b = \frac{k_b}{k_b + k_m}\,P = C\,P,
\qquad C = \frac{k_b}{k_b+k_m},
\]
the \engterm{joint stiffness constant} $C$, while the members shed $(1-C)P$
of their compression. The bolt tension rises only by $C\,P$, not by the full
$P$, because the stiff members absorb most of the change by unloading. The
same split falls out of minimising the total potential energy of the two
parallel springs under the load, the parallel-spring case the
minimum-energy worked example treated.

\textbf{A number.} Take a steel bolt $k_b = 1.0\times 10^{8}\,\si{\newton
\per\meter}$ and a stiffer steel grip $k_m = 4.0\times 10^{8}\,\si{\newton
\per\meter}$, preloaded to $F_i = 20\,\si{\kilo\newton}$, with a service
load $P = 10\,\si{\kilo\newton}$. The joint constant is $C = 1.0/(1.0+4.0)
= 0.20$, so the bolt tension rises by only $\Delta F_b = 0.20\cdot 10 =
2.0\,\si{\kilo\newton}$, to $22\,\si{\kilo\newton}$, while the joint clamp
falls by $8\,\si{\kilo\newton}$ to a residual $12\,\si{\kilo\newton}$,
comfortably still clamped. A fatigue analysis sees the bolt's stress
swinging over only the $2\,\si{\kilo\newton}$ live share, not the full
$10\,\si{\kilo\newton}$, which is why a stiff joint with a soft bolt (a
small $C$) is the fatigue-resistant design and why losing preload, which
opens the joint and hands the bolt the whole load swing, is the failure to
guard against.

\subsection{Worked example: a fixed-fixed beam by symmetry and least work}

A beam built in at both ends carries two redundants past the determinate
count, and least work with the structure's symmetry resolves both with one
short integral. A uniform beam of span $L$ and stiffness $EI$ is clamped at
each end and carries a central point load $W$. Statics gives three
equilibrium equations; the four reactions plus two fixing moments leave the
beam indeterminate to the third degree, but symmetry collapses the unknowns:
the two end reactions are each $W/2$, and the two fixing moments are equal,
call each $M_0$. One redundant, $M_0$, remains.

\textbf{Moment field in terms of the redundant.} Measure $x$ from the left
clamp over the left half, $0\le x\le L/2$. The left reaction $W/2$ acts up
through the arm $x$ and the fixing moment $M_0$ resists, so the bending
moment is
\[
M(x) = \frac{W}{2}\,x - M_0,\qquad 0\le x\le L/2,
\]
and the right half mirrors it. The fixing moment is written as the unknown,
the reaction fixed by symmetry.

\textbf{Least work supplies the missing equation.} The clamp permits no
rotation at the end, so the rotation conjugate to $M_0$ is zero, and
Castigliano reads that rotation as $\partial U/\partial M_0 = 0$. Using the
symmetry to integrate over one half and double,
\[
\frac{\partial U}{\partial M_0}
= 2\int_0^{L/2}\frac{M}{EI}\frac{\partial M}{\partial M_0}\,dx
= \frac{2}{EI}\int_0^{L/2}\!\!\left(\frac{W}{2}x - M_0\right)(-1)\,dx = 0.
\]
The bracket must integrate to zero,
\[
\int_0^{L/2}\!\!\left(\frac{W}{2}x - M_0\right)dx
= \frac{W}{2}\cdot\frac{(L/2)^2}{2} - M_0\,\frac{L}{2}
= \frac{W L^2}{16} - \frac{M_0 L}{2} = 0,
\]
giving
\[
M_0 = \frac{W L}{8}.
\]
The end fixing moments are each $WL/8$, and the moment at the loaded centre
is $M(L/2) = (W/2)(L/2) - WL/8 = WL/8$ as well, equal in magnitude and
opposite in sign to the end moments. The clamps carry half the bending the
load would impose on a simply supported beam of the same span, which is why
fixing the ends of a beam roughly halves its peak moment and stiffens it
against deflection, the structural payoff of a moment connection. The energy
method delivered the redundant fixing moment from one stationarity condition,
with symmetry doing the work that would otherwise need a second
compatibility equation.

\subsection{Worked example: extension and strain energy of a bar under its
own weight}

A bar hanging under its own weight carries an axial force that varies along
its length, and the energy integral handles the variation that a single
$N^2L/(2EA)$ term cannot. A uniform bar of length $L$, area $A$, modulus $E$,
and weight per unit length $w = \rho g A$ hangs from its top end. At a
section a distance $x$ below the top, the bar supports the weight of
everything below it,
\[
N(x) = w\,(L - x),\qquad 0\le x\le L,
\]
the full weight $wL$ at the top tapering linearly to zero at the free bottom.

\textbf{Strain energy.} The axial energy integrates the varying force over
the length,
\[
U = \int_0^L \frac{N(x)^2}{2EA}\,dx
= \frac{w^2}{2EA}\int_0^L (L-x)^2\,dx
= \frac{w^2}{2EA}\cdot\frac{L^3}{3}
= \frac{w^2 L^3}{6EA}.
\]

\textbf{Extension by Castigliano with a tip dummy load.} The total stretch
is the deflection conjugate to a downward load at the free end, so apply a
dummy tip load $Q$, which adds a constant $Q$ to the internal force,
$N(x) = w(L-x) + Q$. Differentiating the energy with respect to $Q$ and
setting $Q=0$,
\[
\delta = \left.\frac{\partial U}{\partial Q}\right|_{Q=0}
= \int_0^L \frac{N}{EA}\frac{\partial N}{\partial Q}\,dx
= \int_0^L \frac{w(L-x)}{EA}\,dx
= \frac{w}{EA}\cdot\frac{L^2}{2}
= \frac{w L^2}{2EA}.
\]
Writing the total weight as $W = wL$, the extension is $\delta = WL/(2EA)$,
exactly half the stretch a load $W$ hung at the tip would produce, because
the self-weight is distributed and most of it acts over only part of the
length. The factor of one half is the signature of a uniformly distributed
axial load, the same factor that distinguishes the $\tfrac12 W\delta$ work
of a gradually applied load from the $W\delta$ of a suddenly applied one.

\textbf{A number.} A steel rod of length $L = 50\,\si{\meter}$ (a deep
mineshaft hoist rope idealised as a bar), area $A = 5.0\times 10^{-4}\,\si
{\meter\squared}$, $E = 200\,\si{\giga\pascal}$, and density $\rho = 7850\,
\si{\kilogram\per\meter\cubed}$ has $w = \rho g A = 7850\cdot 9.81\cdot
5.0\times 10^{-4} \approx 38.5\,\si{\newton\per\meter}$ and total weight
$W = wL \approx 1.93\,\si{\kilo\newton}$. The self-weight extension is
\[
\delta = \frac{W L}{2EA}
= \frac{1925\cdot 50}{2\cdot 200\times 10^{9}\cdot 5.0\times 10^{-4}}
\approx 4.8\times 10^{-4}\,\si{\meter},
\]
about $0.48\,\si{\milli\meter}$, small here but growing with the square of
the length, which is why a kilometre-deep hoist rope must be tapered or
counterweighted: the self-weight stress at the top, $wL/A = \rho g L$, rises
in direct proportion to the depth and sets the limiting length at which the
rope would part under its own weight alone.

\subsection{Worked example: a truss with a lack-of-fit member by virtual
work}

The unit-load method carries an extra term when a member is the wrong
length before assembly, a manufacturing lack of fit, or changes length in
service through a temperature swing. The deflection of a chosen joint is the
sum of the load-induced bar extensions and the imposed lack-of-fit or
thermal extensions, each weighted by the virtual force the unit load puts in
that bar,
\[
\delta = \sum_i n_i\left(\frac{N_i L_i}{E_i A_i} + \lambda_i\right),
\]
with $\lambda_i$ the imposed extension of bar $i$ (positive for a bar made
too long or heated), $N_i$ the real bar force, and $n_i$ the virtual unit-
load bar force. Where the structure is statically determinate the imposed
extensions induce no bar forces at all, since the joints can accommodate
them by moving, and the lack of fit shows up purely as a joint
displacement.

\textbf{The structure.} Reuse the two-bar truss of the earlier unit-load
example: a horizontal bar $BC$ of length $a = 3.0\,\si{\meter}$ and a
diagonal $AC$ of length $L_{AC} = 3.75\,\si{\meter}$ meeting at the loaded
joint $C$, with $\sin\phi = 0.6$, $\cos\phi = 0.8$. The truss carries no
external load now; instead the diagonal $AC$ is fabricated $\Delta =
2.0\,\si{\milli\meter}$ too long. Find the vertical displacement of joint
$C$ that the misfit forces when the bar is pinned home.

\textbf{No real bar forces.} The truss is statically determinate (two bars,
two equilibrium equations at the free joint $C$), so forcing the long
diagonal into place produces no bar forces: the joint simply moves to
absorb the extra length. The real-force term $N_i L_i/(E_i A_i)$ is zero in
every bar, and only the imposed extension of the diagonal survives,
$\lambda_{AC} = +\Delta = +2.0\,\si{\milli\meter}$, with $\lambda_{BC} = 0$.

\textbf{Virtual unit load.} Apply a downward virtual unit load at $C$. From
the earlier statics with $P\to 1$, the virtual bar forces are $n_{AC} =
1/0.6 = 1.667$ and $n_{BC} = -1.333$. The vertical displacement of $C$ is
then
\[
\delta_C = \sum_i n_i\,\lambda_i
= n_{AC}\,\lambda_{AC} + n_{BC}\cdot 0
= 1.667 \cdot 2.0\times 10^{-3}
\approx 3.3\times 10^{-3}\,\si{\meter},
\]
a downward joint movement of about $3.3\,\si{\milli\meter}$ caused entirely
by the oversize diagonal, with no force anywhere in the truss. Replace the
fabrication error by a temperature rise: a steel diagonal warmed by $\Delta
T = 40\,\si{\kelvin}$ extends by $\lambda_{AC} = \alpha L_{AC}\,\Delta T =
12\times 10^{-6}\cdot 3.75\cdot 40 \approx 1.8\times 10^{-3}\,\si{\meter}$,
and the same sum gives $\delta_C = 1.667\cdot 1.8\times 10^{-3} \approx
3.0\,\si{\milli\meter}$ of thermal sag, again with no bar force in the
determinate truss. The lesson is the determinate-versus-indeterminate
split: in a determinate structure imposed strains move joints without
stressing members, while in an indeterminate one the same imposed strains
are resisted and generate self-equilibrating bar forces, which is why
thermal stress is a redundant-structure problem and a determinate truss
expands freely.

\subsection{Worked example: stability of a two-spring snap mechanism}

A mechanism can store its potential energy in a shape whose stationary
points appear and disappear as a load changes, the same snap-through
behaviour the shallow arch showed, in a system simple enough to read by
hand. A rigid rod of length $\ell$ is pinned at its base and held upright by
a horizontal spring of stiffness $k$ attached at the top, the spring
unstretched when the rod is vertical; a vertical load $P$ pushes down on the
top of the rod. We find the load at which the upright rod loses stability.

\textbf{Total potential energy.} Let $\theta$ be the rod's tilt from
vertical. The top of the rod moves horizontally by $\ell\sin\theta$,
stretching the spring, and drops vertically by $\ell(1 - \cos\theta)$,
lowering the load. The total potential energy is the spring energy minus the
load's potential,
\[
\Pi(\theta) = \tfrac12 k (\ell\sin\theta)^2 - P\,\ell(1 - \cos\theta)
= \tfrac12 k \ell^2\sin^2\theta - P\ell(1 - \cos\theta).
\]
The upright position $\theta = 0$ is always an equilibrium, since $\Pi'(0) =
0$ by symmetry. Its stability is the sign of the curvature there.

\textbf{The critical load from the second derivative.} Differentiate twice
and evaluate at $\theta = 0$,
\[
\Pi'(\theta) = k\ell^2\sin\theta\cos\theta - P\ell\sin\theta,
\qquad
\Pi''(0) = k\ell^2 - P\ell.
\]
The upright rod is stable while $\Pi''(0) > 0$, that is while $P < k\ell$,
and loses stability at the critical load
\[
P_\text{cr} = k\ell,
\]
where the curvature of the energy at the upright position passes through
zero. Below $P_\text{cr}$ the upright rod sits in an energy well and returns
when nudged; above it the upright position is an energy maximum and the rod
falls away to one side, the spring no longer stiff enough to hold it. This
is the spring-restrained analogue of Euler buckling, with the discrete
spring stiffness $k$ playing the role the distributed bending stiffness
played for the column, and the critical load reading off the second
variation of a single scalar exactly as the shallow arch and the spring
example did. The two-spring or spring-and-load mechanism is the canonical
teaching model for elastic stability and the simplest place to see a
pitchfork bifurcation of equilibrium emerge from an energy curve.

\begin{mastery}
\textbf{Betti-Maxwell reciprocity and complementary energy.} Two
results deepen the strain-energy machinery. The first is \engterm{Betti's
reciprocal theorem}: for a linearly elastic structure under two separate
load systems $A$ and $B$, the work done by the forces of system $A$
moving through the displacements caused by system $B$ equals the work done
by the forces of system $B$ moving through the displacements caused by
system $A$. Its special case, the \engterm{Maxwell reciprocity} relation,
states that the deflection at point $i$ due to a unit load at point $j$
equals the deflection at $j$ due to a unit load at $i$, so the structure's
flexibility matrix is symmetric. The symmetry is not a convenience of
notation; it is a physical constraint that a correct stiffness or
flexibility matrix must satisfy, and a finite-element matrix that comes
out asymmetric signals a formulation error. The reciprocity follows from
the existence of a strain-energy function that is a state function of the
loads, the same property that makes Castigliano's theorem hold.

The second is the distinction between strain energy $U$, a function of
the displacements, and \engterm{complementary energy} $U^*$, a function
of the loads, related by the Legendre transform $U + U^* = \sum_i P_i
\delta_i$ exactly as the Hamiltonian and Lagrangian are related in
section 5.3. For a linearly elastic structure the force-deflection line
is straight, the two triangular areas are equal, and $U = U^*$ in
magnitude, which is why Castigliano's theorem can be stated against either
energy without confusion. For a nonlinear elastic material the two
energies differ, and the careful statement of Castigliano's first theorem
($P_i = \partial U/\partial\delta_i$, the load as the derivative of strain
energy with respect to displacement) and second theorem ($\delta_i =
\partial U^*/\partial P_i$, the displacement as the derivative of
complementary energy with respect to load) keeps the two apart. The
minimum-potential-energy principle of the spring example is the
displacement-based statement; the principle of minimum complementary
energy is its load-based dual, and the pair is the variational foundation
of the displacement and force methods of structural analysis, and of the
stiffness and flexibility formulations of the finite-element method.
\end{mastery}

\begin{mastery}
\textbf{Least work for many redundants, and influence lines by reciprocity.}
The propped-cantilever worked example resolved a single redundant by setting
$\partial U/\partial R = 0$. A structure with $r$ redundant reactions or
members generalises the idea into the \engterm{force method} of structural
analysis. Choose $r$ redundants $X_1,\ldots,X_r$, release them to leave a
determinate primary structure, and write the internal forces as a
superposition of the actual load case and the $r$ unit-redundant cases,
$M(x) = M_0(x) + \sum_j X_j\,m_j(x)$. The principle of least work demands
that the strain energy be stationary with respect to every redundant
simultaneously, $\partial U/\partial X_i = 0$ for $i = 1,\ldots,r$, which
expands to the linear system
\[
\sum_j \left(\int \frac{m_i\,m_j}{EI}\,dx\right) X_j
= -\int \frac{m_i\,M_0}{EI}\,dx,
\qquad i = 1,\ldots,r,
\]
a symmetric flexibility system $\mathbf{F}\,\mathbf{X} = -\mathbf{d}_0$ whose
coefficient matrix $\mathbf{F}_{ij} = \int m_i m_j/(EI)\,dx$ is the
redundant-to-redundant flexibility and whose right side is the displacement
the load alone would produce at each released redundant. The symmetry
$\mathbf{F}_{ij} = \mathbf{F}_{ji}$ is Maxwell reciprocity again, now as the
structural property that makes the force method's matrix symmetric and
positive definite, so the redundants are the unique solution that minimises
the strain energy. The method scales to any degree of redundancy at the cost
of one $r\times r$ solve, and it is the hand-calculation ancestor of the
flexibility formulation of the matrix structural analysis that finite-element
codes automate.

Reciprocity also delivers \engterm{influence lines}, the diagram of how a
chosen response (a reaction, a bending moment, a deflection) varies as a unit
load tracks across a structure, which the designer needs to place moving
loads for their worst effect. The \engterm{M\"uller-Breslau principle} reads
the influence line off a single deformed shape: the influence line for a
redundant force is, to a scale factor, the deflected shape the structure
takes when a unit displacement is imposed at and in the direction of that
redundant. The principle is a direct consequence of Betti's theorem, with
the moving unit load as one load system and the imposed unit displacement as
the other; the reciprocal-work equality between the two systems turns the
unknown influence ordinate at each load position into the known deflection of
the single released shape. One deformed sketch, drawn by releasing the
response of interest and pushing a unit displacement into it, gives the
entire influence line at once, the load placement that maximises a girder
moment or a support reaction read straight off the shape. The energy
principles thus close the loop from analysis (find the forces in a redundant
structure) to design (place the loads that make those forces worst), both
falling out of the one fact that a linear elastic structure has a symmetric
strain-energy function of its loads.
\end{mastery}

\section{Failure: missing degrees of freedom in dynamic analysis}
\pathtag{core}

The energy method works only as well as the engineer's choice of
generalised coordinates. A coordinate set that fails to capture all
the system's degrees of freedom produces equations of motion that are
correct for the assumed reduced system but blind to motions of the
real system that the coordinates cannot represent. The reduced
analysis predicts the system's response to loads within the
captured modes; the real system's response includes modes the
analysis cannot see. When those uncaptured modes are excited by the
loading, the analysis fails not by computational error but by
missing the question.

The chapter's failure section names the canonical case of this
mode.

\subsection{Tacoma Narrows: a missing torsional degree of freedom}

The original Tacoma Narrows suspension bridge opened on 1 July 1940
and collapsed on 7 November 1940, four months later. On the day of
the collapse, a steady wind of approximately $19\,\si{\meter\per\
second}$ excited a torsional self-excited oscillation of the bridge's
deck; the amplitude grew until structural members failed and the
$853\,\si{\meter}$ centre span tore from its suspension cables and
fell into Puget Sound. The Federal Works Agency report of 28 March
1941, prepared by O.\ H.\ Ammann, Theodore von K\'arm\'an, and Glenn
B.\ Woodruff, identified the collapse as a case of aeroelastic
torsional flutter \cite{acc:tacoma-narrows-1941}.

The mechanism is a coupled motion of the deck and the surrounding
airflow. The deck oscillated in a torsional mode, rotating about its
longitudinal centreline; at each instantaneous angular displacement,
the deck's cross-section presented an angle of attack to the steady
wind, and the wind exerted a pitching moment on the deck that
depended on the angle of attack. At a critical wind speed, the
phase relationship between the deck's angular motion and the
aerodynamic moment became such that the wind did net positive work
on the deck over each cycle, feeding energy into the torsional
oscillation. The amplitude grew exponentially until the structural
capacity of the deck and its connections was exceeded.

The Tacoma Narrows design used the structural-dynamics methods of
its period. The deck was analysed for static wind loads (a horizontal
pressure of $\sim 2.4\,\text{kPa}$ on the projected vertical area)
and for the small-deflection vertical bending modes. The torsional
degree of freedom of the deck section in interaction with the wind
was not part of the analysis. The generalised coordinates of the
structural-dynamics model did not include the deck's torsional mode
coupled to the airflow; the analysis was, in the language of this
chapter, blind to the mode that killed the bridge.

In the analytical-mechanics framework of this chapter, the failure
is precisely a too-small generalised-coordinate set. The deck's
genuine dynamic state requires at least three generalised coordinates
per cross-section: a vertical heave, a horizontal sway, and a
torsional rotation. The aerodynamic loads on the deck depend on all
three; the coupling between deck motion and aerodynamic moment makes
the system not a passive structure under external load but a
self-excited oscillator with an aeroelastic stiffness and aeroelastic
damping. A correctly posed dynamic analysis must include the
torsional coordinate and must include the aerodynamic-feedback term
in the Lagrangian (or equivalently in the equations of motion, as a
generalised non-conservative force coupled to $\dot{\theta}_{
\text{deck}}$). The Tacoma Narrows analysis included neither.

The mechanism reduces to a single equation that the energy method makes
transparent. Model the deck's torsional coordinate $\theta$ as a damped
oscillator driven by an aerodynamic moment that, to leading order,
depends on the torsional rate:
\[
I\ddot{\theta} + c\dot{\theta} + k\theta = M_\text{aero}
\approx q\,A\,\frac{\partial C_M}{\partial\dot\theta}\,\dot\theta,
\]
with $I$ the deck's torsional inertia per unit length, $c$ the
structural damping, $k$ the torsional stiffness, and $q = \tfrac12\rho
U^2$ the dynamic pressure of a wind of speed $U$. Collecting the
rate terms gives an \emph{effective} damping coefficient
\[
c_\text{eff} = c - q\,A\,\frac{\partial C_M}{\partial\dot\theta}.
\]
As long as $c_\text{eff} > 0$ the deck damps disturbances and is stable.
The aerodynamic term grows with the dynamic pressure $q\propto U^2$, so
above a critical wind speed $U_\text{crit}$, where the aerodynamic
contribution overtakes the structural damping, $c_\text{eff}$ goes
negative. A negative damping coefficient is an oscillator that gains
energy each cycle: the amplitude grows as $e^{|c_\text{eff}|t/(2I)}$
until structural members fail. This is the self-excited flutter the
static-load analysis could not see, because the static analysis carried
no $\dot\theta$ term at all and so could not produce a $c_\text{eff}$ to
drive negative. The energy bookkeeping names the failure exactly: the
wind does net positive work on the torsional coordinate per cycle, and
the missing coordinate is the one the work was being done on.

The Federal Works Agency report did not blame Leon Moisseiff, the
bridge's designer (who had also worked on the George Washington and
Golden Gate bridges and was a respected suspension-bridge engineer
of the period); the report identified the deficiency of the
profession's design canon as the underlying mechanism. Suspension-
bridge practice in the 1930s had been moving toward shallower and
more slender deck profiles for aesthetic and economic reasons. The
aerodynamic consequences of those choices had not been brought into
routine design analysis; wind was treated as a static distributed
load. After Tacoma Narrows, suspension-bridge design incorporated
aerodynamic analysis as a standard discipline. The Tacoma Narrows
replacement bridge (1950) used a deeper truss-stiffened deck and
explicit aerodynamic analysis; subsequent suspension-bridge
practice includes wind-tunnel testing of deck section models.

\subsection{The diagnostic principle}

When an energy-method or generalised-coordinate analysis predicts a
result that is contradicted by observation, the first diagnostic
question is whether all the real system's degrees of freedom are
present in the model. The diagnostic check is to enumerate the
modes a complete free-body analysis would represent and to verify
that each appears in the generalised-coordinate set; modes not
represented are silently zeroed by the analysis. The check is
mechanical when the structure is simple (a single rigid body, a
truss) and difficult when the structure is multi-component (a
suspension bridge, a turbine-foundation system, a rocket with
propellant slosh modes). The discipline is the same: count the
degrees of freedom of the real system, then count the generalised
coordinates of the model, then identify the gap.

We return to the discipline in chapter 6 of this volume in the
context of vibration modes (the Tacoma Narrows oscillation was a
specific vibration mode with aerodynamic feedback), in chapter 13
of this volume in the context of bluff-body aerodynamics, and in
chapter 4 of Volume X as the canonical case of an unrecognised
failure mode in an established engineering practice.

\begin{principle}[Generalised coordinates must capture every mode]
An energy-method analysis with $n$ generalised coordinates can
represent at most $n$ independent modes of motion. Modes of the
real system not represented in the coordinate set are zeroed by
the analysis without warning. Before posing a Lagrangian, enumerate
the real system's degrees of freedom and verify that each is in the
coordinate set. The failure mode of an under-coordinated analysis
is not a wrong number; it is a missed phenomenon.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Double pendulum: derive the equations of motion two
ways and reconcile]
The reader derives the equations of motion of a planar double
pendulum first by Newton's method (free-body diagrams of each bob,
explicit tensions in each string, elimination of the tensions),
then by Lagrange's method (kinetic and potential energies in the
swing-angle coordinates, Euler-Lagrange equations). The reader
verifies algebraically that the two derivations produce identical
equations, then implements a numerical simulation of the equations
and explores the motion. The deliverable is a twenty- to forty-page
analysis report.

\textbf{Track}. Derivation plus simulation. \textbf{Hazard class}:
none.

\textbf{Newtonian derivation.} The reader draws free-body diagrams of
each bob; identifies the unknown tensions $T_1$ (upper string) and
$T_2$ (lower string); writes $\vec{F} = m\vec{a}$ for each bob in
appropriate coordinates; eliminates $T_1$ and $T_2$ from the four
component equations to leave two equations of motion in $\theta_1
(t)$ and $\theta_2(t)$.

\textbf{Lagrangian derivation.} The reader writes $T$ and $U$ in the
swing-angle coordinates, forms $L = T - U$, takes the partial
derivatives, and writes out the two Euler-Lagrange equations. The
calculation follows the worked example of section 5.6.2.

\textbf{Reconciliation.} The reader puts the two pairs of equations
side by side and verifies algebraically that they are identical. The
reader writes a one-page commentary on which derivation was faster,
which was less error-prone, and which made the system's structure
(coupling, conservation laws, small-amplitude behaviour) more visible.

\textbf{Simulation.} The reader implements the equations of motion in
a numerical solver (Runge-Kutta or similar) and explores the motion
for at least four sets of initial conditions: (a) both pendula at
small amplitude in phase; (b) both at small amplitude out of phase;
(c) upper at moderate amplitude, lower at rest; (d) both at large
amplitude with energy near the over-the-top regime. For each, the
reader plots $\theta_1(t)$, $\theta_2(t)$, and the system's total
energy as a function of time (which should be constant to within
the integrator's accuracy).

\textbf{Sensitivity test.} For initial conditions (c) or (d), the
reader runs the simulation twice with initial conditions differing
by $10^{-6}$ in $\theta_1(0)$. The reader plots the difference in
$\theta_2(t)$ as a function of time and verifies that the difference
grows exponentially. The exponential growth rate is the Lyapunov
exponent of the double pendulum, a standard signature of
deterministic chaos.

\textbf{Deliverable}. The two derivations side by side, the
reconciliation, the simulation code, the four trajectory plots, the
energy-conservation plot, the sensitivity plot, and a $1{,}500$-word
reflection on what each derivation method made easy or hard to see.

\textbf{Time}. Six to twelve hours of derivation (the Newtonian
derivation is the slow step), four to eight hours of simulation
coding, two to four hours of write-up.
\end{project}

\section{Exercises}

The exercises mix calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment. Solutions are
provided for calculation and derivation exercises; estimation
exercises receive published reference estimates with the editor's
working; design, diagnosis, failure-analysis, and judgment exercises
receive rubrics rather than single answers.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A simple pendulum of length $L = 0.5\,\si{\meter}$ swings with small
amplitude. Find its period and angular frequency.

\paragraph{Solution sketch.} The small-amplitude angular frequency is
$\omega = \sqrt{g/L} = \sqrt{9.81/0.5} \approx 4.43\,\si{\radian\per
\second}$. The period is $T = 2\pi/\omega = 2\pi\sqrt{L/g} \approx
1.42\,\si{\second}$. The result is independent of the bob mass and of
the amplitude, to the small-angle approximation.
\end{exercise}

\begin{exercise}
A bead of mass $m$ slides on a frictionless horizontal wire under no
applied force. State the bead's Lagrangian, write the Euler-Lagrange
equation, and identify the conserved quantity.

\paragraph{Solution sketch.} With $x$ the position along the wire,
$T = \tfrac12 m \dot{x}^2$ and $U = 0$, so $L = \tfrac12 m \dot{x}^2$.
Then $\partial L/\partial\dot{x} = m\dot{x}$ and $\partial L/\partial x
= 0$, so the Euler-Lagrange equation is $\tfrac{d}{dt}(m\dot{x}) = 0$,
hence $m\ddot{x} = 0$. The coordinate $x$ is cyclic, so the conjugate
momentum $p = m\dot{x}$, the linear momentum, is conserved; the bead
moves at constant velocity.
\end{exercise}

\begin{exercise}
A mass $m$ on a spring of stiffness $k$ on a frictionless horizontal
surface has the Lagrangian $L = \tfrac{1}{2} m \dot{x}^2 - \tfrac{1}{2}
k x^2$. Derive the equation of motion. Identify the natural angular
frequency.

\paragraph{Solution sketch.} $\partial L/\partial\dot{x} = m\dot{x}$ and
$\partial L/\partial x = -kx$, so the Euler-Lagrange equation
$\tfrac{d}{dt}(m\dot{x}) + kx = 0$ gives $m\ddot{x} + kx = 0$, that is
$\ddot{x} + (k/m)x = 0$. The natural angular frequency is
$\omega_0 = \sqrt{k/m}$, with period $T = 2\pi\sqrt{m/k}$.
\end{exercise}

\begin{exercise}
A pendulum of length $L$ and mass $m$ is forced by a horizontal force
$F(t) = F_0 \cos\Omega t$ applied to the bob. State the generalised
force $Q^{\text{nc}}$ in the $\theta$-coordinate equation of motion.

\paragraph{Solution sketch.} The generalised force conjugate to
$\theta$ is $Q^{\text{nc}} = \vec{F}\cdot\partial\vec{r}/\partial\theta$.
The bob's position is $\vec{r} = (L\sin\theta,\,-L\cos\theta)$, so
$\partial\vec{r}/\partial\theta = (L\cos\theta,\,L\sin\theta)$. With the
horizontal force $\vec{F} = (F_0\cos\Omega t,\,0)$, the dot product is
$Q^{\text{nc}} = F_0 L \cos\theta\,\cos\Omega t$. The equation of motion
is then $mL^2\ddot{\theta} + mgL\sin\theta = F_0 L\cos\theta\,\cos\Omega
t$; for small $\theta$ this is the forced linear oscillator
$\ddot{\theta} + (g/L)\theta \approx (F_0/mL)\cos\Omega t$.
\end{exercise}

\begin{exercise}
A block of mass $m$ slides on a frictionless inclined plane of angle
$\alpha$. Choose a generalised coordinate, write the Lagrangian,
derive the equation of motion, and find the block's acceleration.

\paragraph{Solution sketch.} Let $s$ be the distance travelled down the
incline. Then $T = \tfrac12 m \dot{s}^2$ and, with the datum at the
start, the height dropped is $s\sin\alpha$, so $U = -mgs\sin\alpha$. The
Lagrangian is $L = \tfrac12 m\dot{s}^2 + mgs\sin\alpha$. The
Euler-Lagrange equation gives $m\ddot{s} - mg\sin\alpha = 0$, hence the
acceleration $\ddot{s} = g\sin\alpha$, independent of mass and the
familiar incline result.
\end{exercise}

\begin{exercise}
A bead of mass $m$ slides on a frictionless rotating hoop of radius
$R$ rotating at $\Omega = 2\sqrt{g/R}$. Find the bead's stable
equilibrium angles on the hoop.

\paragraph{Solution sketch.} The off-axis equilibria satisfy
$\cos\theta_c = g/(R\Omega^2)$. With $\Omega^2 = 4g/R$, $R\Omega^2 = 4g$,
so $\cos\theta_c = g/(4g) = 1/4$, giving $\theta_c \approx \pm 75.5^\circ$.
Since $R\Omega^2 = 4g > g$, the bottom $\theta = 0$ is unstable and the
two off-axis points $\pm 75.5^\circ$ are the stable equilibria. (The
top $\theta = \pi$ is always unstable.) This matches the
$\Omega/\Omega_\text{crit} = 2$ row of \path{data/hoop_equilibria.csv}.
\end{exercise}

\begin{exercise}
A double pendulum has equal masses $m$ and equal string lengths $L$.
Write the small-amplitude (linearised) equations of motion. Find the
two natural frequencies of small-amplitude oscillation.

\paragraph{Solution sketch.} Linearising the equal-mass equal-length
double pendulum gives $2\ddot{\theta}_1 + \ddot{\theta}_2 +
2(g/L)\theta_1 = 0$ and $\ddot{\theta}_1 + \ddot{\theta}_2 +
(g/L)\theta_2 = 0$. Substituting $\theta_i = a_i\cos\omega t$ yields the
eigenvalue problem $(K - \omega^2 M)a = 0$ with $M =
\bigl(\begin{smallmatrix}2&1\\1&1\end{smallmatrix}\bigr)$ and
$K = (g/L)\bigl(\begin{smallmatrix}2&0\\0&1\end{smallmatrix}\bigr)$. The
determinant condition gives $\omega^2 = (2\mp\sqrt2)\,g/L$, so
$\omega_1 = \sqrt{(2-\sqrt2)g/L} \approx 0.765\sqrt{g/L}$ (in-phase) and
$\omega_2 = \sqrt{(2+\sqrt2)g/L} \approx 1.848\sqrt{g/L}$ (out-of-phase),
matching \path{data/double_pendulum_modes.csv}.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the Euler-Lagrange equation from Hamilton's principle.
State the boundary-condition assumption and identify where it is
used in the integration-by-parts step.

\paragraph{Solution sketch.} Vary the action $S = \int_{t_1}^{t_2}
L(q,\dot q,t)\,dt$ by $q\to q+\delta q$ with $\delta q(t_1) = \delta
q(t_2) = 0$ (the fixed-endpoint assumption). To first order, $\delta S =
\int [\,(\partial L/\partial q)\delta q + (\partial L/\partial\dot
q)\delta\dot q\,]\,dt$. Integrate the second term by parts:
$\int (\partial L/\partial\dot q)\,\delta\dot q\,dt = [(\partial
L/\partial\dot q)\delta q]_{t_1}^{t_2} - \int \tfrac{d}{dt}(\partial
L/\partial\dot q)\,\delta q\,dt$. The boundary term vanishes precisely
because $\delta q$ is zero at both endpoints; this is where the
boundary-condition assumption is used. Then $\delta S = \int [\,\partial
L/\partial q - \tfrac{d}{dt}(\partial L/\partial\dot q)\,]\delta q\,dt$,
and requiring $\delta S = 0$ for arbitrary $\delta q$ forces the
bracket to vanish, giving $\tfrac{d}{dt}(\partial L/\partial\dot q) -
\partial L/\partial q = 0$.
\end{exercise}

\begin{exercise}
Show that the Euler-Lagrange equation for a Lagrangian $L = \tfrac{1}{2}
m \dot{q}^2 - U(q)$ in a single generalised coordinate $q$ reduces to
Newton's second law for a particle moving in the potential $U$.

\paragraph{Solution sketch.} Here $\partial L/\partial\dot q = m\dot q$,
so $\tfrac{d}{dt}(\partial L/\partial\dot q) = m\ddot q$, and $\partial
L/\partial q = -dU/dq$. The Euler-Lagrange equation $\tfrac{d}{dt}
(\partial L/\partial\dot q) - \partial L/\partial q = 0$ becomes $m\ddot
q + dU/dq = 0$, that is $m\ddot q = -dU/dq = F(q)$. This is Newton's
second law for a particle of mass $m$ in the potential $U$, with the
conservative force $F = -dU/dq$.
\end{exercise}

\begin{exercise}
A particle moves in a central force field with potential $U(r)$.
Choose polar coordinates $(r, \phi)$. Write the Lagrangian. Identify
the cyclic coordinate and the corresponding conserved momentum. State
the physical meaning of the conserved momentum.

\paragraph{Solution sketch.} In polar coordinates the velocity squared
is $\dot r^2 + r^2\dot\phi^2$, so $T = \tfrac12 m(\dot r^2 +
r^2\dot\phi^2)$ and $L = \tfrac12 m(\dot r^2 + r^2\dot\phi^2) - U(r)$.
The coordinate $\phi$ does not appear in $L$ (only $\dot\phi$ does), so
$\phi$ is cyclic and $p_\phi = \partial L/\partial\dot\phi =
mr^2\dot\phi$ is conserved. This is the angular momentum about the
force centre; its conservation is the rotational symmetry of the central
potential, equivalent to Kepler's law of equal areas in equal times.
\end{exercise}

\begin{exercise}
Derive the small-amplitude (linearised) equations of motion of a
double pendulum with equal masses and equal string lengths. Identify
the two normal modes and their natural frequencies.

\paragraph{Solution sketch.} Linearising as in the calculation-block
companion gives $M = \bigl(\begin{smallmatrix}2&1\\1&1\end{smallmatrix}
\bigr)$, $K = (g/L)\bigl(\begin{smallmatrix}2&0\\0&1\end{smallmatrix}
\bigr)$. The eigenvectors of $M^{-1}K$ are the mode shapes. The
low-frequency mode $\omega_1^2 = (2-\sqrt2)g/L$ has the bobs in phase
with amplitude ratio $a_2/a_1 = +\sqrt2/2$; the high-frequency mode
$\omega_2^2 = (2+\sqrt2)g/L$ has them out of phase with $a_2/a_1 =
-\sqrt2$. The lower link swings with larger amplitude than the upper in
both modes. General small-amplitude motion is a superposition of the
two, each oscillating at its own frequency; the resulting beat pattern
is what makes a lightly disturbed double pendulum appear to trade energy
between the links.
\end{exercise}

\begin{exercise}
Derive the bifurcation condition $R\Omega^2 = g$ for the bead-on-
rotating-hoop system of section 5.6.3. State the physical
interpretation of the bifurcation.

\paragraph{Solution sketch.} The equilibrium condition is
$\sin\theta\,(g - R\Omega^2\cos\theta) = 0$. The off-axis solutions
$\cos\theta_c = g/(R\Omega^2)$ require the right side to lie in
$(0,1)$, which is possible only when $R\Omega^2 > g$. As $\Omega$ rises
through the critical value $\Omega_\text{crit} = \sqrt{g/R}$, the
condition $R\Omega^2 = g$ marks the appearance of the two new
equilibria, simultaneously with the bottom losing stability. The
interpretation: below the threshold the gravitational restoring effect
dominates and the bead rests at the bottom; above it the centrifugal
effect of the rotation exceeds gravity at the bottom, the bottom becomes
unstable, and the bead is flung outward to a new stable angle. The
transition is a pitchfork bifurcation, a symmetry-breaking event in
which the single symmetric equilibrium gives way to a symmetric pair.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the period of a $1\,\si{\meter}$ pendulum swinging at
amplitude $\theta_0 = 45^{\circ}$. State the correction to the small-
amplitude formula and bracket the answer.

\paragraph{Reference estimate.} Small-angle period $T_0 = 2\pi\sqrt{1/
9.81} \approx 2.006\,\si{\second}$. With $\theta_0 = 45^\circ = 0.785\,
\si{\radian}$, the first correction $T \approx T_0(1 + \theta_0^2/16) =
2.006\,(1 + 0.0386) \approx 2.083\,\si{\second}$. The exact value from
\path{data/pendulum_period.csv} is $2.086\,\si{\second}$, so the true
period sits between the small-angle $2.01\,\si{\second}$ and roughly
$2.09\,\si{\second}$, about $4\,\%$ above the small-angle figure. Bracket:
$2.01$ to $2.09\,\si{\second}$.
\end{exercise}

\begin{exercise}
Estimate the energy stored in the elastic deformation of a typical
diving board when supporting a $70\,\si{\kilogram}$ diver at full
deflection. State assumptions about board geometry and material.

\paragraph{Reference estimate.} Treat the board as a cantilever
deflecting by $\delta \approx 0.15\,\si{\meter}$ under the diver's
weight $W = mg \approx 686\,\si{\newton}$ at the tip. A linear-spring
model stores $U = \tfrac12 W\delta$ at full static deflection (the
factor of one-half because the force builds up linearly with the
deflection), so $U \approx \tfrac12 \cdot 686 \cdot 0.15 \approx
51\,\si{\joule}$. Order of magnitude: tens of joules, comparable to the
$mg\,h$ a diver gains by rising $\sim 0.1\,\si{\meter}$. Assumptions: a
springboard tip deflection of order $0.1$ to $0.2\,\si{\meter}$ and a
roughly linear load-deflection relation up to full deflection. A
fibreglass board returning most of this energy is what gives the
spring its launch.
\end{exercise}

\begin{exercise}
Estimate the critical rotation speed at which a bead on a $30\,\si
{\milli\meter}$-radius rotating hoop transitions from the bottom-
equilibrium regime to the off-axis-equilibrium regime. Express in
revolutions per minute.

\paragraph{Reference estimate.} The threshold is $R\Omega^2 = g$, so
$\Omega_\text{crit} = \sqrt{g/R} = \sqrt{9.81/0.030} \approx
18.1\,\si{\radian\per\second}$. Converting, $f = \Omega/(2\pi) \approx
2.88$ revolutions per second, or about $173$ revolutions per minute.
This matches the value printed by \path{code/hoop_bifurcation.py}.
Below roughly $170$ rev/min the bead sits at the bottom; above it the
bead climbs to an off-axis equilibrium.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a numerical Runge-Kutta solver for the simple-pendulum
equation $\ddot{\theta} + (g/L)\sin\theta = 0$. Compare the simulated
period to the small-amplitude prediction for swing amplitudes
$10^{\circ}$, $30^{\circ}$, $60^{\circ}$, and $90^{\circ}$. Plot the
period as a function of amplitude.

\paragraph{Solution sketch.} The reference implementation is
\path{code/pendulum_period.py}: integrate $\ddot\theta + (g/L)\sin\theta
= 0$ as a first-order system from rest at $\theta_0$, and measure the
period from the time to return to the turning point (or four times the
time to the first downward zero crossing). Expected results, from
\path{data/pendulum_period.csv} with $L = 1\,\si{\meter}$, $T_0 \approx
2.006\,\si{\second}$: at $10^\circ$ the period exceeds $T_0$ by about
$0.19\,\%$; at $30^\circ$ by $1.7\,\%$; at $60^\circ$ by $7.3\,\%$; at
$90^\circ$ by $18\,\%$. The period rises monotonically with amplitude
and diverges logarithmically as $\theta_0\to 180^\circ$. The first
correction $1 + \theta_0^2/16$ tracks the curve to better than $0.5\,\%$
out to about $60^\circ$.
\end{exercise}

\begin{exercise}
Implement a Lagrangian-mechanics solver for the double pendulum.
Verify energy conservation to within $0.1\,\%$ over a $30\,\si{\second}$
simulation. Compare the trajectories for nearby initial conditions
and demonstrate exponential divergence (the chapter's project takes
this further).

\paragraph{Solution sketch.} The reference implementation is
\path{code/double_pendulum.py}: form the two-by-two mass matrix from the
linearised-acceleration coefficients, invert it at each step to extract
$(\ddot\theta_1,\ddot\theta_2)$, advance with RK4, and log the total
energy $E = T + U$ at each step. A correctly implemented conservative
integrator with a step of $10^{-3}\,\si{\second}$ holds $E$ constant to
under $2\times 10^{-7}$ relative over $30\,\si{\second}$, comfortably
inside the $0.1\,\%$ target, as recorded in
\path{data/double_pendulum_trajectory.csv}. Repeating a large-amplitude
run with $\theta_1(0)$ perturbed by $10^{-6}\,\si{\radian}$ produces a
separation $|\Delta\theta_2(t)|$ that grows exponentially, a straight
line on a log scale (figure~\ref{fig:vol03ch05:double-pendulum-divergence}),
until it saturates near one radian. Energy drift growing faster than
this bound signals a coding error in the mass-matrix inversion or too
large a step.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a pendulum clock with period exactly $2\,\si{\second}$.
Specify the pendulum length and the regulating-mechanism principle.
State the temperature-induced length change that would shift the
clock by $1\,\si{\second\per\day}$.

\paragraph{Model-answer sketch.} For $T = 2\,\si{\second}$, the
small-amplitude relation $L = g(T/2\pi)^2$ gives $L = 9.81\cdot
(1/\pi)^2 \approx 0.994\,\si{\meter}$, the classic ``seconds pendulum''
of just under a metre (its half-period, one swing, is one second). The
regulating mechanism is an escapement that delivers a small impulse each
swing to replace damping losses, keeping the amplitude small and
constant so the isochronism of the small-angle period holds. Sensitivity:
$T\propto\sqrt L$, so $\Delta T/T = \tfrac12\,\Delta L/L$. A drift of
$1\,\si{\second}$ per day is $\Delta T/T = 1/86400 \approx
1.16\times 10^{-5}$, hence $\Delta L/L = 2.3\times 10^{-5}$, or $\Delta
L \approx 23\,\si{\micro\meter}$ on a one-metre rod. A steel rod
($\alpha \approx 12\times 10^{-6}\,\si{\per\kelvin}$) reaches this with
a temperature change of about $2\,\si{\kelvin}$, which is why precision
clocks use temperature-compensated gridiron or invar pendulums.
\end{exercise}

\begin{exercise}
Design a torsion pendulum (rotor on a thin wire) to measure small
torques. Specify the rotor's moment of inertia, the wire's torsion
stiffness, and the natural period of the pendulum. State the minimum
torque the pendulum can detect, given a $1$-arcsecond angular
resolution.

\paragraph{Model-answer sketch.} The torsion pendulum obeys $I\ddot
\phi + \kappa\phi = \tau_\text{ext}$, with $I$ the rotor moment of
inertia and $\kappa$ the wire torsional stiffness; the natural period is
$T = 2\pi\sqrt{I/\kappa}$. A worked choice: a rotor of $I \approx
10^{-4}\,\si{\kilogram\meter\squared}$ on a wire of $\kappa \approx
10^{-6}\,\si{\newton\meter\per\radian}$ gives $T = 2\pi\sqrt{10^{-4}/
10^{-6}} = 2\pi\cdot 10 \approx 63\,\si{\second}$, a slow, easily read
oscillation. The detectable torque follows from the static deflection
$\tau = \kappa\,\phi_\text{min}$: with $\phi_\text{min} =
1\,\text{arcsec} = 4.85\times 10^{-6}\,\si{\radian}$, the minimum torque
is $\tau \approx 10^{-6}\cdot 4.85\times 10^{-6} \approx 5\times
10^{-12}\,\si{\newton\meter}$. The design trade is explicit: a softer
wire (smaller $\kappa$) detects smaller torques but lengthens the period
and raises sensitivity to drift and seismic noise. This is the
Cavendish-balance principle.
\end{exercise}

\begin{exercise}
Design a vibration test rig for evaluating the natural frequency of
a small cantilever beam. State the generalised coordinates required,
the energy expressions, and the predicted natural frequency for a
$300\,\si{\milli\meter}$ aluminum beam of $1\,\si{\milli\meter}$
thickness.

\paragraph{Model-answer sketch.} One generalised coordinate suffices
for the fundamental mode: the tip deflection $w(t)$, with an assumed
mode shape $\phi(x)$ satisfying the clamped-free boundary conditions.
Kinetic energy $T = \tfrac12 m_\text{eff}\dot w^2$ and elastic potential
$U = \tfrac12 k_\text{eff} w^2$ give $\omega_1 = \sqrt{k_\text{eff}/
m_\text{eff}}$. The exact Euler-Bernoulli result is $\omega_1 =
(1.875)^2\sqrt{EI/(\rho A L^4)}$. For an aluminium beam ($E \approx
69\,\si{\giga\pascal}$, $\rho \approx 2700\,\si{\kilogram\per\meter
\cubed}$) of width $b$, thickness $h = 1\,\si{\milli\meter}$, length
$L = 0.3\,\si{\meter}$, the width cancels: $I/A = h^2/12$, so $\omega_1 =
3.516\sqrt{E h^2/(12\rho L^4)} = 3.516\sqrt{69\times 10^9\cdot 10^{-6}/
(12\cdot 2700\cdot 0.3^4)} \approx 3.516\cdot 8.2 \approx
29\,\si{\radian\per\second}$, that is $f_1 \approx 4.6\,\si{\hertz}$. The
rig drives the beam base near this frequency and detects the resonant
amplitude peak. A Rayleigh estimate with a parabolic trial shape lands
within a few per cent of this.
\end{exercise}

\begin{exercise}
Design a Foucault pendulum installation in a building atrium to
demonstrate Earth's rotation. Specify the pendulum length, the
swing amplitude, the expected precession rate at the installation's
latitude, and the indicator-pin spacing on the demonstration
fixture.

\paragraph{Model-answer sketch.} A long pendulum keeps the swing plane
visibly fixed while the building rotates beneath it. Choose $L \approx
20\,\si{\meter}$ for a period $T = 2\pi\sqrt{20/9.81} \approx
9\,\si{\second}$ and a heavy bob to fight air damping; swing amplitude a
few degrees. The precession rate of the swing plane is $\Omega_p =
\Omega_\oplus\sin\lambda$, with $\Omega_\oplus = 2\pi$ per sidereal day
$= 15.04^\circ$ per hour and $\lambda$ the latitude. At $\lambda =
45^\circ$, $\Omega_p \approx 10.6^\circ$ per hour, a full turn in about
$34\,\si{\hour}$. Over one swing of amplitude $A \approx 2\,\si{\meter}$
at the floor, the plane advances by $A\,\Omega_p\,T \approx 2\cdot
(10.6\cdot\pi/180/3600)\cdot 9 \approx 9\times 10^{-4}\,\si{\meter}$, so
indicator pins spaced a few millimetres apart are knocked over one per
many minutes, a readable demonstration. State the installation latitude
explicitly, since the rate vanishes at the equator and is maximal at the
poles.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A flexible-joint robot arm is analysed using rigid-link approximations
and exhibits unexpected oscillations at the end-effector under fast
moves. Identify the missing degree of freedom and state the modeling
correction.

\paragraph{Solution sketch.} The rigid-link model assigns one
generalised coordinate per joint and treats each link as
infinitely stiff. The missing degree of freedom is the elastic
deflection of the joint (and of the link), a compliance the rigid model
zeroes. Under slow moves the joint stiffness keeps the deflection
negligible; under fast moves the inertial torque excites the
joint-compliance mode, and the end-effector rings at the natural
frequency $\sqrt{k_\text{joint}/I_\text{link}}$. The correction is to add
a coordinate for each flexible joint, modelling it as a torsional spring
$k_\text{joint}$ in series with the actuator, giving a flexible-joint
Lagrangian with twice as many coordinates as the rigid model. The
controller must then either be bandwidth-limited below the compliance
mode or include explicit vibration damping.
\end{exercise}

\begin{exercise}
A car's suspension is analysed as a two-degree-of-freedom system
(sprung mass and unsprung mass on a tire spring). The car exhibits an
unexplained vibration mode in the $5\,\si{\hertz}$ range. Identify
the most probable missing degree of freedom and state the
diagnostic measurement.

\paragraph{Solution sketch.} The two-degree-of-freedom quarter-car model
captures the sprung-mass (body) bounce near $1\,\si{\hertz}$ and the
unsprung-mass (wheel) hop near $10$ to $15\,\si{\hertz}$. A mode near
$5\,\si{\hertz}$ falls between these and is most probably a structural
flexible mode the lumped model omits: body torsional or bending
flexibility, a subframe resonance, or seat-and-occupant dynamics. The
missing degree of freedom is the chassis or subframe elastic mode, a
coordinate the rigid-body two-mass model cannot represent. Diagnostic
measurement: an instrumented modal test (accelerometers along the body
with a shaker or impact hammer) to extract the mode shape at
$5\,\si{\hertz}$; if the shape is a chassis bending or subframe motion
rather than rigid body-on-suspension motion, the lumped model needs the
extra coordinate.
\end{exercise}

\begin{exercise}
A rocket's attitude control uses a model that treats the propellant
as a rigid mass at the centre of the tank. The control loop exhibits
limit cycles when the tank is partially full. Identify the missing
degree of freedom and state the engineering correction.

\paragraph{Solution sketch.} Treating the propellant as a rigid mass
zeroes the sloshing of the free liquid surface. The missing degree of
freedom is the propellant slosh mode: in a partially full tank the
liquid surface can rock, and to first order this behaves like a pendulum
(or spring-mass) of natural frequency set by the tank radius and fill
level. The slosh mass couples to the vehicle's attitude through the
tank-wall reaction, and when the slosh frequency lies near the control
loop's bandwidth the feedback sustains a limit cycle. The correction is
to add a slosh coordinate, modelled as an equivalent pendulum or
spring-mass attached to the vehicle, into the attitude-dynamics
Lagrangian, and to add baffles inside the tank that raise the slosh
damping and shift its frequency away from the control bandwidth. This is
the classic propellant-slosh problem of launch-vehicle guidance.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
Re-read the Federal Works Agency report on the Tacoma Narrows Bridge
collapse \cite{acc:tacoma-narrows-1941}. Identify the generalised
coordinates that a corrected aeroelastic analysis would include and
state which aerodynamic feedback term in the Lagrangian (or equivalent
generalised force) drove the instability.

\paragraph{Solution sketch.} A corrected aeroelastic model needs, per
deck cross-section, at least three generalised coordinates: vertical
heave $h$, horizontal sway $s$, and torsional rotation $\alpha$ about the
deck's longitudinal axis. The original analysis carried only the
vertical bending coordinate under static wind load and omitted $\alpha$.
The destabilising term is the aerodynamic pitching moment that depends on
the torsional rate $\dot\alpha$, entering as a generalised
non-conservative force $Q_\alpha = \partial M_\text{aero}/\partial\dot
\alpha\cdot\dot\alpha$ whose sign over a cycle is such that the wind does
net positive work on the deck. This negative aerodynamic damping (an
effective damping coefficient driven below zero) makes the torsional
oscillation self-excited: the amplitude grows exponentially rather than
decays. The report's diagnosis of aeroelastic torsional flutter is
exactly the statement that the $\alpha$ coordinate and its
aerodynamic-feedback term were absent from the design model
\cite{acc:tacoma-narrows-1941}.
\end{exercise}

\begin{exercise}
A wind-loaded tall structure (chimney, antenna mast, slender tower)
exhibits unexpected vortex-induced oscillation at moderate wind
speeds. State, in $200$ to $300$ words, the generalised coordinates
that a complete analysis must include and the aerodynamic-coupling
term that drives the response.

\paragraph{Model-answer sketch.} A complete analysis carries the
cross-wind transverse displacement of the structure (and, for tall
towers, the first few bending-mode coordinates) rather than treating
wind as a static drag load. The driving mechanism is vortex shedding:
the alternating wake vortices apply a transverse force at the Strouhal
frequency $f_s = St\,U/D$, with $St \approx 0.2$ for a circular section,
$U$ the wind speed, $D$ the diameter. When $f_s$ approaches a structural
natural frequency, the oscillation and the shedding lock in: the
structure's own motion organises the wake, so the forcing term becomes a
function of the response (a motion-dependent generalised force), and the
amplitude can grow until limited by nonlinearity. The coordinate that
matters is the transverse modal displacement; the coupling term is the
lift force synchronised to that displacement during lock-in. A complete
model adds aerodynamic damping that can go negative over the lock-in
range. A strong answer names lock-in explicitly and distinguishes it from
ordinary forced resonance.
\end{exercise}

\begin{exercise}
A bridge cable-stay system, designed for static wind loads only,
exhibits significant cable galloping in icing conditions. State the
missing degree of freedom in the static-load analysis and the
mechanism by which icing changes the aerodynamic-coupling term.

\paragraph{Solution sketch.} The static-load analysis omits the cable's
transverse oscillation coordinate (and, where relevant, its torsional
rotation), treating wind as a fixed drag. Galloping is a one-degree-of-
freedom aeroelastic instability of the transverse coordinate. A bare
cable is nearly circular and aerodynamically stable; an ice accretion
gives the cross-section a non-circular profile (a D-shape or crescent)
whose lift coefficient varies steeply with angle of attack. The Den
Hartog galloping criterion is that instability occurs when $dC_L/d\alpha
+ C_D < 0$, that is when the slope of the lift curve plus the drag turns
the effective aerodynamic damping of the transverse coordinate negative.
Icing changes the coupling term by reshaping the section so that
$dC_L/d\alpha$ becomes large and negative; the wind then feeds energy
into the transverse mode and the cable gallops at large, low-frequency
amplitude. The correction is aerodynamic (de-icing, section shaping) or
structural (added damping, cross-ties between cables).
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A junior engineer asks whether Lagrangian mechanics replaces Newtonian
mechanics or supplements it. Write a one-paragraph reply that
addresses when each method is preferred and when they are
interchangeable.

\paragraph{Model-answer sketch.} Lagrangian mechanics supplements
rather than replaces Newtonian mechanics: the two are mathematically
equivalent, derivable from each other through d'Alembert's principle, so
they always give the same equations of motion. The choice is tactical.
Newton's method is preferred when the constraint and contact forces are
themselves wanted (bearing loads, joint reactions, tensions), since
those forces are eliminated in the Lagrangian formulation and must be
recovered separately. The Lagrangian method is preferred for systems
with many degrees of freedom, complicated holonomic constraints, or
non-Cartesian coordinates, where free-body bookkeeping is error-prone and
the scalar-energy route is cleaner. For a single particle or a simple
rigid body the two are interchangeable and the engineer picks whichever
is faster. A strong reply notes that the methods are one physics in two
notations.
\end{exercise}

\begin{exercise}
A senior colleague argues that energy methods are an academic
luxury and that working engineers should use free-body diagrams
exclusively. Write a one-paragraph reply that takes the argument
seriously and states the design situations in which energy methods
are essential.

\paragraph{Model-answer sketch.} The colleague is right that for many
everyday problems, a single rigid body, a statically determinate
structure, a simple machine element, a free-body diagram is the fastest
and clearest tool, and the senior engineer's fluency with it is
valuable. The argument fails as a general policy. Energy methods become
essential when the system has many degrees of freedom and many internal
constraints (multi-link robots, vehicle multibody dynamics), where
free-body bookkeeping grows quadratically and the unknown internal
forces swamp the calculation; when the analysis is variational by nature
(finite-element structural dynamics, Rayleigh-Ritz natural-frequency
estimates, optimal-control trajectory problems); and when the failure
mode is a missing degree of freedom, where the discipline of writing one
scalar energy and enumerating coordinates is the safeguard the
Tacoma Narrows designers lacked. A strong reply concedes the practical
point and bounds it: free-body diagrams for the simple and the
local, energy methods for the many-coordinate and the variational.
\end{exercise}

\begin{exercise}
Identify a mechanical system in your immediate environment whose
dynamics you have never analysed in the Lagrangian formulation
(spring-loaded door, oscillating fan, vibrating washing machine).
State, in $200$ to $300$ words, the generalised coordinates you
would choose, the kinetic and potential energy expressions, and
whether the system has any cyclic coordinates that imply a
conservation law.

\paragraph{Model-answer sketch.} The answer is open, but a strong one
follows the chapter's procedure for whatever system is chosen. Take a
spring-loaded screen door: one generalised coordinate, the opening angle
$\theta$; kinetic energy $T = \tfrac12 I\dot\theta^2$ with $I$ the
door's moment of inertia about the hinge; potential energy combining the
closer-spring term $\tfrac12 k_\theta\theta^2$ and, if the door is not
vertical-axis, a gravity term. The damper makes the system
non-conservative, so the generalised damping force enters as
$Q = -c\dot\theta$. There is no cyclic coordinate here, since $\theta$
appears in $U$, so no conservation law; the energy decays through the
damper. A reply that instead picks a system with a genuine symmetry (a
freely spinning fan coasting down, where the spin angle is cyclic and
angular momentum is conserved until friction acts) should identify the
cyclic coordinate and the conserved momentum explicitly. The graded
skill is correct identification of coordinates, energies, dissipation,
and any symmetry, not the particular object.
\end{exercise}
