\chapter{Population biology, ecology, epidemiology}
\label{vol06:ch08}

\epigraph{All models are wrong, but some are useful.}{George E. P.
Box, in a section of the 1979 collection on the analysis and design
of experiments \cite[p.~202]{paper:v6c08-box-1979}.}

\chapmeta{Half-life: durable for the underlying mathematics; specific
cohorts and datasets are medium-life. Archetypes: scaling, transport,
control. Exercise target: 30. Project track: simulation, with one
optional household-scale fermentation analogue. Hazard class: low (no
pathogen handling; all field epidemiology is via published datasets).
Reader path default: standard, with sections explicitly tagged. Named
cases: SARS-CoV-2 modelling failures (no single registry entry; cited
from peer-reviewed retrospectives).}

\noindent
A population is a balance: births minus deaths, immigration minus
emigration, infection minus recovery. The same accounting that
governs heat in a reactor or mass in a pipeline governs the size of
a bacterial colony, a deer herd, a city, an epidemic wave, and a
species on the edge of extinction. This chapter develops the
balance for populations as the engineer reads it: as a set of
coupled ordinary differential equations with measurable parameters,
observable failure modes, and finite predictive horizon. The reader
who finishes the chapter can fit an SIR model to a published outbreak
dataset, estimate when a fishery will collapse from its harvest
record, write down the resistance trajectory of an antibiotic-treated
population, and identify which of the model's assumptions broke when
its prediction missed.

\section{Population dynamics: growth and carrying capacity}\pathtag{core}

The simplest population model is a single state variable $N(t)$, the
number of individuals at time $t$, with a single rate law. If
individuals reproduce at per-capita rate $r$ independently of
density, the equation is
\[
\frac{dN}{dt} = r N,
\]
and the solution is exponential, $N(t) = N_0 e^{r t}$. The
per-capita rate $r$ has units of inverse time and is the difference
between the per-capita birth rate $b$ and the per-capita death rate
$d$. For human populations in demographic transition, $r$ runs from
$-0.005$ (Japan, current as of 2024) to $+0.030$
per year (sub-Saharan Africa, current as of 2024); for a bacterial
culture in log phase, $r$ is of order $1\,\text{h}^{-1}$, six orders
of magnitude faster than human demographics. The model fails on
either side of its short validity window: birth and death rates
themselves depend on density, on environment, on the age distribution
of the population, and on prior history.

\engterm{The logistic equation}. The first refinement caps the
exponential by a density-dependent term. The Belgian mathematician
Pierre-Fran\c{c}ois Verhulst introduced the model in 1838
\cite[pp.~117--121]{hist:v6c08-verhulst-1838}, motivated by Quetelet's data on
nineteenth-century population growth in Belgium:
\[
\frac{dN}{dt} = r N \left(1 - \frac{N}{K}\right),
\]
where $K$ is the \engterm{carrying capacity}, the equilibrium
population the environment supports. The solution is the logistic
curve,
\[
N(t) = \frac{K}{1 + \left(\frac{K - N_0}{N_0}\right) e^{-r t}},
\]
which approaches $K$ as $t \to \infty$ from any positive initial
condition. The inflection point is at $N = K/2$, where the growth
rate is maximised at $r K/4$. The carrying capacity is the
phenomenological aggregate of the limits that constrain growth:
food, water, shelter, predation, disease, waste accumulation. The
logistic equation does not say what $K$ is; it parameterises the
limit.

\engterm{Verhulst's derivation, in modern form}. The reasoning runs
through three steps. (1) Exponential growth $dN/dt = rN$ is the
limit of a per-capita rate that does not depend on density. (2) In
finite environments, the per-capita rate must vanish at some
limit $K$ and be negative above it. (3) The simplest analytic form
that meets both ends is linear in $N$: the per-capita rate is
$r(1 - N/K)$. Substituting gives the logistic equation. Separating
variables yields
\[
\int \frac{dN}{N(1 - N/K)} = \int r\,dt,
\]
and partial-fraction decomposition $\frac{1}{N(1 - N/K)} =
\frac{1}{N} + \frac{1/K}{1 - N/K}$ produces $\ln N - \ln(1 - N/K) =
rt + C$. Solving for $N$ and evaluating $C$ from the initial
condition $N(0) = N_0$ gives the closed-form solution above. The
derivation is the canonical separable ODE; the population biology is
in the choice of the linear correction, not in the integration.

\engterm{Maximum sustainable yield and the logistic spine}. A
harvested logistic population with constant harvest rate $H$ is
governed by $dN/dt = rN(1 - N/K) - H$. Equilibria solve
$H = rN(1 - N/K)$. The maximum equilibrium harvest is at $N = K/2$
and equals $H_{\max} = rK/4$, the \engterm{maximum sustainable
yield} (MSY). Any harvest level $H < H_{\max}$ admits two equilibria:
a stable high-population equilibrium and an unstable low-population
equilibrium. Reading the curve from the practitioner's side: a
fishery managed at exactly $H_{\max}$ sits on a knife-edge. Any
recruitment shortfall, mis-estimate of $r$ or $K$, or random
mortality pulse drops the population below $K/2$, at which point the
harvest exceeds production and the population collapses
deterministically. The 1970s North-Atlantic-fishery MSY orthodoxy
took the calculation as a target rather than as a ceiling; a
generation later, the same calculation reads as the predictable cause
of the cod collapse (§8.7). The engineering correction is to set
harvest at a fraction of MSY (commonly $2/3$ to $4/5$), to use the
buffer to absorb estimation error and stochastic variation, and to
treat $K$ as something to be measured directly rather than inferred
from current biomass.

\input{volumes/06-life/ch08-population-ecology-epidemiology/figures/fig-logistic-growth.tex}

\engterm{Empirical fit and its limits}. The logistic model fits
laboratory cultures of yeast, paramecia, and \emph{Drosophila} well
\cite[ch.~3]{text:v6c08-gotelli-ecology}. It fits some natural
populations (sheep in Tasmania, reindeer on St.\ Matthew Island) well
for a few generations and then misses badly. The failures are
informative. Yeast in a flask experience a real carrying capacity:
the medium contains a finite amount of glucose, and the yeast
metabolise it to a finite limit. Sheep on a finite island experience
a more complex limit: when the population approaches the carrying
capacity, soil erosion, vegetation succession, and parasite load all
shift, so $K$ itself moves. A model with a fixed $K$ predicts
oscillation around equilibrium; a model with a $K$ that drifts
predicts overshoot and collapse. The Tasmanian sheep population in
the nineteenth century overshot, the vegetation degraded, and the
sheep crashed; the reindeer on St.\ Matthew Island in 1944--1966 grew
from 29 to 6000 and then crashed to 42 in a single winter
\cite[pp.~350--367]{paper:v6c08-klein-1968-reindeer}.

\engterm{Time-delay logistic}. When the carrying capacity response
lags, a delay-differential model captures the overshoot:
\[
\frac{dN}{dt} = r N(t) \left(1 - \frac{N(t - \tau)}{K}\right).
\]
For $r \tau < \pi/2$, the system damps toward $K$; for larger
$r \tau$, sustained oscillations emerge. The delay $\tau$ is set by
generation time, by the regeneration time of the resource, or by the
time scale over which density-dependent effects feed back to birth or
death rates. For deer browsing on forest understory, $\tau$ is a
decade or more; for plankton blooms responding to nutrient depletion,
$\tau$ is days. Volume IX develops the engineering of delay systems
in control terms; here the same dynamics arise as a property of the
ecological feedback loop.

\begin{archetype}[Scaling: from the per-capita rate to the population]
The per-capita rate $r$ is a property of individuals: how often they
reproduce, how often they die. The population is its integral over
the cohort. The logistic equation is the first nonlinear
correction, in which the per-capita rate $r$ becomes a function of the
density $N/K$. Every subsequent ecological model can be read as a
further refinement of the per-capita rate: as a function of
predator density, of prey density, of climate, of infection
prevalence, of age. The reader who tracks how the per-capita rate
becomes denser owns the engineering structure of population biology.
\end{archetype}

\engterm{Age-structured populations}. Single-state models pool
individuals of all ages. Real populations have age structure, and
births depend on the number of reproductive adults rather than on
the total. The Leslie matrix \cite[ch.~5]{text:v6c08-caswell-matrix-models}
is the discrete-time, age-structured generalisation. For a population
with $n$ age classes, the state vector $\mathbf{n}_t = (n_{1,t},
n_{2,t}, \ldots, n_{n,t})^\top$ evolves by
\[
\mathbf{n}_{t+1} = L \mathbf{n}_t,
\]
where $L$ is the Leslie matrix with fertilities $F_i$ on the top row
and survival probabilities $P_i$ on the subdiagonal. The dominant
eigenvalue $\lambda$ of $L$ is the asymptotic growth rate per time
step ($\lambda > 1$ means growth, $\lambda < 1$ decline), and the
corresponding right eigenvector is the \engterm{stable age
distribution}. The reader who computes $\lambda$ from a Leslie
matrix gets the same intelligence the demographer gets from a life
table.

\engterm{Allee effects}. At low density, some populations grow
slower than the per-capita rate predicts, because mate-finding,
cooperative defence, or social facilitation requires a minimum density
to be effective. The \engterm{strong Allee effect} is captured by
\[
\frac{dN}{dt} = r N \left(\frac{N}{A} - 1\right)\left(1 - \frac{N}{K}\right),
\]
with $A$ the Allee threshold below which the population shrinks even
in a favourable environment. Conservation biology takes the Allee
threshold seriously: a population reduced below $A$ is in a basin of
attraction that does not include recovery, and a one-off intervention
that raises the population above $A$ may be the difference between
extinction and persistence. The black-footed ferret programme in the
United States used captive breeding to push the wild population
above its empirical Allee threshold; the kakapo programme in New
Zealand has done the same for a flightless parrot whose breeding
biology requires a critical density of fertile males. The engineering
question is the threshold's location; the engineering action is
delivering the population to a state above it.

Courchamp, Clutton-Brock, and Grenfell catalogued the field's
evidence for Allee effects in 1999 \cite[pp.~405--410]{paper:v6c08-courchamp-allee-1999}.
Their list separates cooperative-defence Allee effects (meerkats,
African wild dogs), mate-finding Allee effects (passenger pigeons in
the late nineteenth century, when local densities fell below the
threshold for sustained colonial nesting), and predator-satiation
Allee effects (sea-bird colonies whose collective breeding swamps
egg predators only above a threshold colony size). The black-footed
ferret population fell to 18 individuals in 1985; captive breeding
raised the wild population back through 50 by 1991 and through 1000
by 2008, and the threshold language was load-bearing in the funding
case. The engineering reading: an Allee threshold is a discontinuity
in the population's basin of attraction, and the action that delivers
the population through it is what the model identifies.

\begin{mastery}
The continuous logistic equation is well-behaved: smooth approach
to $K$ for any positive initial condition, no oscillation, no
chaos. Its discrete-time analogue is a different animal. May's
1976 \emph{Nature} paper \cite[pp.~459--467]{paper:v6c08-may-1976-chaos}
analysed
\[
N_{t+1} = r N_t \left(1 - \frac{N_t}{K}\right),
\]
where the same arithmetic now produces a one-dimensional map. For
$r < 1$, the population dies out. For $1 < r < 3$, the
population converges to a stable fixed point. For $3 < r < 1 +
\sqrt{6} \approx 3.449$, the fixed point loses stability and the
trajectory settles on a stable two-cycle. Further increases in $r$
produce a four-cycle, an eight-cycle, and so on; the period-doubling
cascade accumulates at the Feigenbaum constant
$\delta \approx 4.669$, after which the dynamics are chaotic with
embedded windows of periodicity. The result reframed the
diversity-stability debate: a one-species model with a single
parameter can produce dynamics indistinguishable from random noise.
The reader who fits a logistic-style discrete-time model to a real
population should plot the bifurcation diagram for the fitted $r$
before claiming the residuals reflect environmental stochasticity;
some residuals are the model's deterministic chaos.
\end{mastery}

\begin{estimation}
\textbf{Question.} A laboratory \emph{Escherichia coli} culture starts
with $10^4$ cells per millilitre in a 100\,mL flask. The cells divide
every 30 minutes in log phase. The carrying capacity of the medium is
$\sim 10^9$ cells/mL. Estimate the time to reach $5 \times 10^8$
cells/mL.

\smallskip
\textit{Estimate before reading on.}

\smallskip
The doubling time is 30 minutes, so $r = \ln 2 / 0.5\,\text{h} =
1.39\,\text{h}^{-1}$. Initial density $N_0 = 10^4$/mL; target density
$N = 5 \times 10^8$/mL; the target is half the carrying capacity,
just below where logistic correction begins to bite. Pure
exponential growth would reach the target in
\[
t = \frac{1}{r} \ln \frac{N}{N_0} = \frac{1}{1.39} \ln (5 \times 10^4)
\approx \frac{10.8}{1.39} \approx 7.8\,\text{h}.
\]

\smallskip
\textit{Postmortem.} The logistic correction at $N = K/2$ slows
growth by a factor of two; integrating the logistic equation gives
$t \approx 8.0\,\text{h}$, within 3\% of the exponential estimate.
The estimate would fail badly if the target were $0.9 K$, where
logistic deceleration dominates and exponential timing
underestimates by a factor of three. A working bench biologist who
knows the doubling time and the carrying capacity can read off
elapsed time to mid-log phase with this kind of arithmetic; the more
common failure is forgetting that the culture itself depletes oxygen,
ages, and changes its $r$ over the course of the run.
\end{estimation}

\section{Predator-prey models}\pathtag{standard}

Two coupled state variables: prey $N(t)$ and predator $P(t)$. The
prey grow in the absence of predators; the predators die in the
absence of prey; predators eat prey, killing prey and feeding their
own reproduction. Alfred Lotka in 1925 and Vito Volterra in 1926
independently wrote the canonical equations
\cite[pp.~88--110]{hist:v6c08-lotka-1925, hist:v6c08-volterra-1926}:
\[
\frac{dN}{dt} = \alpha N - \beta N P, \qquad
\frac{dP}{dt} = \delta N P - \gamma P,
\]
with $\alpha$ the prey birth rate, $\gamma$ the predator death rate,
$\beta$ the predation rate constant, and $\delta = \epsilon \beta$
the conversion of consumed prey into predator biomass, $\epsilon$
the conversion efficiency. The equations have a non-trivial
equilibrium at $N^* = \gamma/\delta$, $P^* = \alpha/\beta$. Around
this equilibrium the system oscillates, with period $T = 2\pi/
\sqrt{\alpha \gamma}$ and amplitude set by the initial condition.
The oscillations are neutrally stable: there is no damping, no
limit cycle, and a small perturbation produces a different
amplitude. Volterra developed the equations to explain why the
fraction of predator fish landed in Adriatic ports rose during
1914--1918 when fishing was suppressed: less fishing means more
prey, which means more predators, which means a higher predator
fraction. His brother-in-law D'Ancona, the fisheries biologist, had
seen the data; Volterra wrote the model.

\input{volumes/06-life/ch08-population-ecology-epidemiology/figures/fig-lotka-volterra-phase.tex}

\engterm{Phase-portrait analysis}. Reading the Lotka-Volterra
equations as a planar dynamical system, the prey nullcline ($dN/dt
= 0$) is the line $P = \alpha/\beta$, and the predator nullcline
($dP/dt = 0$) is the line $N = \gamma/\delta$. The two lines cross
at the equilibrium $(N^*, P^*)$. The flow on either side of each
nullcline is perpendicular to the nullcline in question, so the
trajectories rotate counterclockwise around the equilibrium. The
Jacobian at $(N^*, P^*)$ is
\[
J^* = \begin{pmatrix} 0 & -\beta N^* \\ \delta P^* & 0 \end{pmatrix},
\]
with eigenvalues $\pm i\sqrt{\alpha\gamma}$. Both eigenvalues lie on
the imaginary axis; the equilibrium is a centre rather than a focus,
and small perturbations produce closed orbits rather than spirals.
The orbits are level curves of the conserved quantity
\[
H(N, P) = \delta N - \gamma \ln N + \beta P - \alpha \ln P,
\]
which can be checked by computing $dH/dt$ and substituting the
equations. The conservation law is exact for the deterministic
Lotka-Volterra system; the neutral stability is a structural
prediction of the model. Adding a logistic prey term $-rN^2/K$ to
the prey equation breaks the conservation law and converts the
centre into a stable focus, with damped oscillations toward
equilibrium. This is the simplest path from Lotka-Volterra to the
Rosenzweig-MacArthur model below.

\engterm{Limits of Lotka-Volterra}. The model is a teaching tool,
not a quantitative predictor. The neutral stability is unphysical: a
real population perturbed off its limit cycle does not stay on a
new one. The Rosenzweig-MacArthur model
\cite[pp.~209--223]{paper:v6c08-rosenzweig-macarthur-1963} adds a logistic
prey, a saturating predator functional response (Holling type II,
$\beta N/(N + h)$ instead of $\beta N$), and converts the neutral
centre into a true equilibrium that the system approaches with
damped oscillation or, in some parameter regimes, a stable limit
cycle. The Lynx-Hare cycles documented in the Hudson's Bay Company
records of 1845--1935 (with a period of about 10 years) are the
canonical empirical case; modern fits use a Rosenzweig-MacArthur
type model with additional environmental forcing
\cite[ch.~5]{text:v6c08-may-stability}.

\engterm{Paradox of enrichment}. The Rosenzweig-MacArthur model
\[
\frac{dN}{dt} = rN\left(1 - \frac{N}{K_N}\right) - \frac{\beta N P}{N + h},
\qquad
\frac{dP}{dt} = \frac{\delta N P}{N + h} - \gamma P,
\]
has a non-trivial equilibrium at $N^* = \gamma h / (\delta - \gamma)$
and a corresponding $P^*$ that depends on $K_N$. Linear-stability
analysis at the equilibrium reveals a Hopf bifurcation as $K_N$
crosses a critical value $K_N^{\mathrm{crit}}$: below it, the
equilibrium is stable and the predator-prey system damps to
equilibrium; above it, a stable limit cycle takes over and the
populations oscillate at finite amplitude. The implication is
counterintuitive in the agronomic frame: improving the prey's habitat
(raising $K_N$) destabilises the predator population, producing
amplitude oscillations that increase the probability of stochastic
extinction. This is the \engterm{paradox of enrichment}, the most
discussed result in theoretical ecology of the 1970s. The engineering
reading is precise: the system's qualitative behaviour depends on
a parameter that the ecosystem manager controls (nutrient input,
habitat quality), and the relationship is not monotone in the
species the manager intends to protect.

\engterm{Functional responses}. Holling classified the predator
response to prey density into three types. Type I is linear
($\beta N$): the predator's consumption rate scales with prey
density without saturation, valid only for filter feeders or
herbivores with effectively unlimited handling capacity. Type II is
hyperbolic ($\beta N/(N + h)$): the predator saturates as prey
become abundant because handling time and digestion impose an
upper limit. The handling-time interpretation gives $h$ the meaning
of the prey density at which the predator achieves half its
maximum consumption rate. Type III is sigmoid ($\beta N^2/(N^2 +
h^2)$): the predator's consumption rises slowly at low prey density,
either because the predator switches to alternative prey or because
the predator has trouble locating sparse prey. Each functional
response changes the stability of the model in characteristic
ways; the type III response is the principal mechanism by which
sparse prey populations can be stabilised against extinction by a
predator that switches to alternative prey.

\engterm{Multi-species food webs}. A real ecosystem has more than
two species. Lawton, May, and others developed the theory of food
webs in the 1970s and 1980s
\cite[chs.~6--7]{text:v6c08-may-stability}. A food-web matrix records who
eats whom, weighted by interaction strength. The web's stability is
determined by the eigenvalues of the community matrix; May's theorem
states that, for randomly chosen interactions, the probability of
stability decreases with the size of the web, in apparent
contradiction to the empirical observation that large ecosystems
are stable. The resolution is that real ecosystems have structured
interactions: most links are weak, the strong links are arranged
in patterns that confer stability (compartmentalisation, weak omnivory,
adaptive foraging), and the structure is the product of selection
on the food web itself. The engineering analogy is to a large
sparse matrix versus a dense one: the matrix's stability follows
from its sparsity and from the location of its non-zero entries,
not from the entries' values alone.

\section{SIR and SEIR epidemic models}\pathtag{core}

A communicable disease spreads through a population by contact
between infected and susceptible individuals. The simplest model,
introduced by Kermack and McKendrick in 1927
\cite[pp.~700--721]{paper:v6c08-kermack-mckendrick-1927}, partitions the
population into three compartments: \engterm{susceptible} ($S$),
\engterm{infectious} ($I$), and \engterm{recovered} ($R$). The
equations are
\begin{align*}
\frac{dS}{dt} &= -\beta S I / N, \\
\frac{dI}{dt} &= \beta S I / N - \gamma I, \\
\frac{dR}{dt} &= \gamma I,
\end{align*}
with $\beta$ the contact-and-transmission rate per susceptible per
infectious per unit time (per total population), $\gamma$ the
recovery rate (inverse of mean infectious period $D = 1/\gamma$), and
$N = S + I + R$ the constant total population. The model assumes
homogeneous mixing (every susceptible is equally likely to contact
every infectious), a fixed infectious period, and permanent immunity
on recovery.

\input{volumes/06-life/ch08-population-ecology-epidemiology/figures/fig-sir-flow.tex}

\engterm{The basic reproduction number $R_0$}. The expected number of
secondary infections caused by one infectious individual introduced
into a fully susceptible population. From the SIR equations,
\[
R_0 = \frac{\beta}{\gamma}.
\]
The derivation, in the form the modern literature uses, runs through
the early-phase linearisation. At $t = 0$ a single infectious
individual is introduced into a fully susceptible population, so
$S \approx N$. The infectious-compartment equation becomes
$dI/dt = (\beta - \gamma) I$, with solution $I(t) = I_0
e^{(\beta - \gamma)t}$. The exponential growth rate is
$\lambda = \beta - \gamma$. Each infectious individual remains
infectious for an average duration $D = 1/\gamma$, during which
the individual contacts $\beta D = \beta/\gamma$ susceptibles and
infects each. The product is $R_0 = \beta/\gamma$, the
\engterm{next-generation matrix} reduction of a one-compartment
infectious model. For models with more than one infectious
compartment (SEIR with an exposed class, multi-strain models,
age-stratified models), $R_0$ is the dominant eigenvalue of the
next-generation matrix $FV^{-1}$, where $F$ is the matrix of new
infections and $V$ the matrix of compartment transitions. The
formula above is the scalar special case.
$R_0$ is the most-cited single number in epidemiology. For measles,
$R_0$ is in the range 12--18 (the highest among common human
pathogens); for influenza, 1.3--2.0; for SARS-CoV-1, 2--3; for the
ancestral SARS-CoV-2 strain, the consensus estimate from 2020--2021
meta-analyses is 2.5--3.0 \cite[Tab.~1]{paper:v6c08-liu-covid-r0-2020}; for
the Omicron variant of late 2021, much higher, with most estimates in
the range 6--10. Each of these numbers is conditional on the
population's contact structure at the time of measurement;
$R_0$ is a property of the pathogen-in-the-population, not of the
pathogen alone.

\input{volumes/06-life/ch08-population-ecology-epidemiology/figures/fig-r0-regions.tex}

\engterm{The herd-immunity threshold}. If a fraction $1 - 1/R_0$ of
the population is immune (by infection or vaccination), the
effective reproduction number $R = R_0 S/N$ drops below 1, and an
introduced infection dies out. The derivation is short: at the
threshold, $R = 1$, so the immune fraction satisfies $R_0 (1 - p_c)
= 1$, which rearranges to $p_c = 1 - 1/R_0$. For measles, the threshold is
$\approx 0.94$; for the ancestral SARS-CoV-2, $\approx 0.65$; for
Omicron, $\approx 0.90$. The threshold is a calculation, not a goal:
herd immunity is the steady state in which $R = 1$, and a fraction
of the originally susceptible population is unavoidably infected
along the way to that state. The fraction infected by the end of an
unmitigated epidemic is the larger root of the final-size relation
\[
1 - s_\infty = 1 - e^{-R_0 (1 - s_\infty)},
\]
where $s_\infty$ is the fraction never infected. For $R_0 = 2.5$,
$s_\infty \approx 0.107$, so about 89\% of the population is
eventually infected; for $R_0 = 6$, $s_\infty \approx 0.0025$, so
about 99.7\% is eventually infected. The herd-immunity threshold
and the final-size fraction are different quantities, often
conflated.

\engterm{The SEIR model}. Many pathogens have a latent period during
which an infected individual is not yet infectious. The SEIR model
adds an \engterm{exposed} compartment $E$:
\begin{align*}
\frac{dS}{dt} &= -\beta S I / N, \\
\frac{dE}{dt} &= \beta S I / N - \sigma E, \\
\frac{dI}{dt} &= \sigma E - \gamma I, \\
\frac{dR}{dt} &= \gamma I,
\end{align*}
with $\sigma$ the inverse latent period. The basic reproduction number
is still $R_0 = \beta/\gamma$, unchanged by the latent compartment,
but the dynamics on a finite time horizon differ: the latent period
delays the epidemic's peak, lowers the early growth rate, and
matters for the design of interventions that work on the latent state
(quarantine of exposed contacts, for example).

\engterm{Compartments beyond SEIR}. Real-world modelling almost
always extends the basic structure. SEIRD adds a dead compartment $D$
with a case-fatality rate $\mu$. SEIRS allows reinfection with rate
$\omega$ (waning immunity), appropriate for influenza and SARS-CoV-2.
Age-structured models stratify by age band, each with its own
parameters and its own contact matrix to the others. Spatial models
add a location index. Stochastic models replace the deterministic ODE
with a continuous-time Markov chain, important when the number of
infectious individuals is small (early-outbreak phase, end-of-outbreak
phase, low-prevalence endemic state). The engineer who models an
outbreak picks the structure that captures the questions the
intervention needs answered.

\begin{principle}[Compartmental models partition the population by disease state]
A compartmental epidemic model is an accounting equation for a
population stratified by disease state. Each compartment represents
an indistinguishable group of individuals; transitions between
compartments are governed by per-capita rates that may depend on
the prevalences of other compartments. The model's predictive power
comes from the conservation of total population, the explicit
dependence of the transmission rate on the susceptible-infectious
contact, and the constraints from observed time series. The model's
limits come from the assumption that the stratification is exhaustive
and that the per-capita rates do not depend on individual history,
network position, or behaviour.
\end{principle}

\engterm{Fitting an SIR to data}. The standard practice is to
estimate $R_0$ from the early-phase exponential growth rate. If
$I(t) \approx I_0 e^{\lambda t}$ during the early phase (when
$S \approx N$), then $\lambda \approx \beta - \gamma$ and $R_0
\approx 1 + \lambda/\gamma$. If $\gamma$ is known from clinical
data on infectious period and $\lambda$ is measured from incidence
time series, $R_0$ follows. Maximum-likelihood and Bayesian
methods refine the estimate using the full time series and
incorporating noise; the practical engineering check is
whether the predicted peak time, peak height, and final size
match the observed time series, and whether the residual
errors are consistent with the model's stochastic assumptions.
The Wallinga-Teunis 2004 estimator
\cite[pp.~509--516]{paper:v6c08-wallinga-teunis-2004} extends the
exponential-growth method to a time-resolved estimator of $R_t$ from
incidence and serial-interval data; the calculation underwrites most
public dashboards of the effective reproduction number during the
2020--2023 SARS-CoV-2 pandemic. Age-stratified models input a
contact matrix, of which the POLYMOD survey
\cite[pp.~e74-1--e74-15]{paper:v6c08-mossong-polymod-2008} provides
the standard European reference.

\begin{estimation}
\textbf{Question.} The cumulative confirmed SARS-CoV-2 case count
in a country doubles every 2.5 days during the first three weeks of
its first wave. The mean infectious period from the early clinical
literature is 5 days. Estimate $R_0$ and the herd-immunity
threshold.

\smallskip
\textit{Estimate before reading on.}

\smallskip
The doubling time $T_d = 2.5\,\text{d}$ gives an exponential growth
rate $\lambda = \ln 2 / T_d = 0.28\,\text{d}^{-1}$. The recovery
rate $\gamma = 1/D = 0.20\,\text{d}^{-1}$. The basic reproduction
number is then
\[
R_0 \approx 1 + \frac{\lambda}{\gamma} = 1 + \frac{0.28}{0.20}
= 2.4,
\]
inside the 2.5--3.0 consensus range for the ancestral SARS-CoV-2
strain reported by the Liu et al.\ 2020 meta-review
\cite[Tab.~1]{paper:v6c08-liu-covid-r0-2020}. The herd-immunity
threshold is $p_c = 1 - 1/R_0 \approx 0.58$.

\smallskip
\textit{Postmortem.} The doubling-time formula assumes a
homogeneously mixing fully susceptible population, an SIR-compatible
single exponential infectious period, and case ascertainment that
does not change with the wave. The first assumption shifts $R_0$
downward (heterogeneity inflates the early growth rate because
high-contact individuals are infected first). The second assumption
underestimates $R_0$ for SEIR-type pathogens with a latent period,
because the latent compartment delays the exponential growth and
the formula above attributes the slow growth to a low $R_0$ rather
than to a long generation time. The third assumption shifts both
ways: ramping up case ascertainment during the wave inflates
$\lambda$; saturating testing capacity at peak deflates it. The
working engineering response is to treat the doubling-time estimate
as a lower-bound calculation and to widen the uncertainty band by
a factor of 1.5--2 before publishing.
\end{estimation}

\section{Ecosystems as networks}\pathtag{standard}

An ecosystem is a network: nodes are species (or guilds, or
functional groups), edges are interactions (predation, competition,
mutualism, parasitism). The network has a topology, an edge weight
distribution, and a dynamics. Engineering analysis of an ecosystem
is the analysis of the network's structure and its consequences.

\engterm{Food-web topology}. Real food webs are sparse: species
interact with a small fraction of the other species in the web. The
connectance $C$, defined as the fraction of possible directed pairs
that interact, is typically 0.05 to 0.30 across observed food webs
\cite[ch.~4]{text:v6c08-may-stability}. The webs are nearly hierarchical:
basal species (plants, algae, detritus) at the bottom, primary
consumers above, predators at the top, with omnivory weaker than
strict hierarchy would predict. The trophic levels are not integer
in practice; a species' trophic position is the weighted average
of its prey's positions plus one.

\input{volumes/06-life/ch08-population-ecology-epidemiology/figures/fig-food-web.tex}

\engterm{Productivity, biomass, and trophic transfer}. Eugene
Odum's energy-flow analysis of ecosystems
\cite[chs.~3--4]{text:v6c08-odum-1971} computes the gross
primary productivity (GPP) at the base of the food web (the rate
at which photosynthetic organisms fix carbon), the net primary
productivity (NPP $=$ GPP minus autotroph respiration), and the
fraction of NPP that reaches each subsequent trophic level. The
classical ``10\% rule'' (each trophic level retains roughly 10\% of
the energy entering it from the level below) is a rule of thumb;
the empirical transfer efficiency is closer to 2\%--20\% across
aquatic and terrestrial systems. The rule explains why food chains
are short (5--6 trophic levels typically, never more than 7--8): the
energy budget runs out.

\engterm{Nutrient cycling}. Carbon, nitrogen, phosphorus, and other
biogeochemical cycles thread through the food web with their own
network topology. Carbon flows from atmospheric CO$_2$ through
primary producers, through consumers, back to the atmosphere as
respiration and to soils as detritus. Nitrogen is fixed from
atmospheric N$_2$ by a small set of bacteria and cyanobacteria,
incorporated into amino acids and nucleic acids, and returned to
the atmosphere by denitrifying bacteria. The engineer who wants to
manipulate an ecosystem (a wetland for water treatment, a farm soil
for nitrogen retention, a coral reef for resilience) does so by
modifying the nutrient-cycle network.

\engterm{Connectivity, robustness, and the power-law tail}. The
distribution of links per species in a food web is heavy-tailed:
most species have few connections, a small number of generalist
species have many. Albert, Jeong, and Barab{\'a}si's 2000
\emph{Nature} paper \cite[pp.~378--382]{paper:v6c08-albert-barabasi-2000}
established the engineering reading of scale-free networks (degree
distribution $P(k) \sim k^{-\alpha}$ with $\alpha$ in the 2--3
range): such networks are extremely robust to random node removal
(retaining global connectivity even when 80\% of nodes are removed)
and extremely fragile to targeted attack on high-degree nodes
(connectivity collapses after removal of a few percent of hubs).
Dunne, Williams, and Martinez applied the framework to empirical
food webs \cite[pp.~558--567]{paper:v6c08-dunne-foodweb-2002}: the
food-web analogue of the targeted-attack failure is the loss of
highly connected ``hub'' species, which drag through a cascade of
secondary extinctions. The keystone-species concept introduced by
Paine in his 1966 intertidal-experiment paper
\cite[pp.~65--75]{paper:v6c08-paine-1966} is the network-topological
shorthand for the same phenomenon: removing one species of starfish
restructured the community by eliminating predation on dominant
competitors. The engineering reading is precise: an ecosystem's
robustness is not a single number but a function on the species
list, and conservation prioritisation that protects hubs delivers
more robustness per dollar than uniform protection.

\engterm{Resilience and tipping points}. An ecosystem in a stable
state remains in that state under small perturbations. A large
perturbation, or a slow shift in external forcing past a critical
threshold, can produce an alternative stable state. Lake eutrophication
is the canonical example: a shallow lake with low phosphorus loading
is clear-water, vegetated, and biologically diverse; the same lake
with high phosphorus loading is turbid, algae-dominated, and low
diversity. The transition between the two states is hysteretic: once
the lake has flipped to the turbid state, reducing phosphorus to the
loading level that previously sustained the clear state does not
restore the clear state, because internal phosphorus cycling and
self-shading of vegetation have changed
\cite[ch.~5]{text:v6c08-scheffer-shifts}. The engineer who manages an
ecosystem watches for early-warning signals (rising variance, slowing
recovery after perturbation, increased spatial correlation) and
intervenes before the tipping point is crossed.

\section{Antibiotic resistance: the evolution of resistance}\pathtag{core}

A bacterial population exposed to an antibiotic faces a selection
gradient: cells with mutations conferring resistance survive and
reproduce; cells without such mutations die. The dynamics are the
same as natural selection in general, but the time scale is short
(generations of hours), the selection coefficient is large (death
of susceptible cells, fitness of resistant cells often higher than
the susceptible wild type in the presence of drug), and the
mutational supply is enormous (a single overnight culture of
$10^9$ cells contains, by mutation rate $\sim 10^{-9}$ per nucleotide
per generation, on average one cell with any given single
point mutation). Resistance is not a possible outcome of antibiotic
use; it is the inevitable outcome unless the conditions of use
suppress it.

\engterm{Mechanisms of resistance}. Bacterial resistance to
antibiotics arises through four mechanisms
\cite[chs.~2--3]{text:v6c08-walsh-resistance}: enzymatic inactivation of
the drug ($\beta$-lactamases against penicillins and cephalosporins,
aminoglycoside-modifying enzymes); modification of the drug's
target (alteration of the penicillin-binding proteins, ribosomal
methylation against macrolides); reduced drug uptake (membrane
permeability changes); active efflux (transporters that pump the
drug out of the cell faster than it diffuses in). Resistance genes
spread vertically by clonal expansion and horizontally by conjugative
plasmids, transposons, integrons, and bacteriophages. The horizontal
transfer is a feature that distinguishes bacterial evolution from
eukaryotic evolution and accelerates resistance spread across
species boundaries.

\engterm{Fitness costs and compensation}. Resistance often imposes a
fitness cost in the absence of the drug: the resistant cell grows
more slowly, or competes less effectively for nutrients. The cost
predicts that resistance frequencies fall when drug pressure is
relaxed, and the prediction holds for some resistances (penicillin
resistance in certain pneumococcal lineages declined in Iceland
when prescribing was reduced). For other resistances, compensatory
mutations restore the fitness deficit while preserving resistance;
once compensated, the resistant lineage no longer pays a cost and
no longer recedes under drug withdrawal. The engineering question is
the time scale of compensation; the working answer is that
compensation is often quick, often within a year of drug withdrawal
in the laboratory.

\engterm{Population-genetic models of resistance}. Let $f$ be the
fraction of resistant cells in a population. In the presence of drug
at concentration $c$ above the inhibitory concentration for
susceptible cells, the resistant cells grow with rate $r_R$ and the
susceptible with rate $-d_S$ (death). The resistant fraction grows by
\[
\frac{df}{dt} = (r_R + d_S) f (1 - f),
\]
the logistic equation in the resistant fraction. The fixation time
(the time for $f$ to rise from $f_0$ to $1 - f_0$) is
\[
t_{\text{fix}} = \frac{2}{r_R + d_S} \ln \frac{1 - f_0}{f_0}.
\]
For a starting fraction $f_0 = 10^{-9}$ (one resistant cell in a
billion susceptible) and a selection differential $r_R + d_S = 1\,
\text{h}^{-1}$, the fixation time is about $20 \cdot 2 = 41\,\text{h}$,
under two days. The numbers map directly to the clinical observation
that monotherapy with a drug to which resistance can arise by a
single mutation often fails within a few days to a few weeks. The
operational response is combination therapy (multiple drugs requiring
multiple independent mutations) and the engineering of dosing
regimens that maintain drug concentrations above the
mutant-selection window.

\engterm{The mutant-selection window}. At very low drug concentration,
neither susceptible nor resistant cells are inhibited and no
selection occurs. At very high concentration, both are inhibited and
no growth occurs. Between these limits is a range, the
\engterm{mutant-selection window}, within which susceptible cells
are inhibited but resistant cells are not. The window is the locus
of selection for resistance. Dosing regimens that minimise time
spent in the window (high-peak, short-duration; combination with
synergistic drugs; cycling) are the engineering response.

\input{volumes/06-life/ch08-population-ecology-epidemiology/figures/fig-resistance-timeline.tex}

\engterm{Selection coefficient and the time scale of fixation}. In
population-genetic notation, the selection coefficient $s$ is the
fractional fitness advantage of the resistant allele relative to the
susceptible: $s = (r_R - r_S)/r_S$ where $r_R$ and $r_S$ are the
per-generation growth rates. The time for a beneficial allele to
sweep from frequency $f_0$ to fixation is approximately
\[
t_{\mathrm{fix}} \approx \frac{1}{s} \ln \frac{1 - f_0}{f_0}
\]
generations for $s \ll 1$, and shorter when $s$ is large. For an
antibiotic at therapeutic concentration the selection differential is
of order unity per generation (susceptible cells die rather than grow
slowly), so the fixation time is on the order of $\ln(f_0^{-1})$
generations regardless of the precise value of $s$. With $f_0 =
10^{-9}$, the figure is $\sim 21$ generations; with $f_0 = 10^{-6}$
(plasmid-mediated resistance pre-existing in the patient
microbiota), $\sim 14$ generations. Two days is a generous upper
bound. Fisher's 1937 wave-of-advance result
\cite[pp.~355--369]{paper:v6c08-fisher-1937} provides the spatial
extension: in a diffusing population, the resistant front advances at
constant speed $v = 2\sqrt{D r_R}$ where $D$ is the diffusion
coefficient of the bacterial population. The wave-of-advance
calculation underpins the temporal structure of hospital outbreaks of
resistance, in which a resistant strain identified in one ward
appears at predictable speed across the institution.

The fitness-cost-and-compensation literature
\cite[pp.~260--271]{paper:v6c08-andersson-hughes-2010} adds the
realistic correction: $s$ is positive in the presence of drug and
often negative (or modestly positive after compensation) in its
absence. The clinical implication is that stopping antibiotic use
returns the resistant frequency toward $f_0$ slowly when no
compensation has occurred, but does not return at all when
compensatory mutations have fixed the cost. The global health
budget reads in this arithmetic: Murray et al.\ 2022 in \emph{The
Lancet} \cite[pp.~629--655]{paper:v6c08-murray-amr-2022} estimated
1.27 million deaths directly attributable to bacterial AMR in 2019,
with 4.95 million associated deaths, more than HIV/AIDS or malaria
that year.

\begin{estimation}
\textbf{Question.} A new antibiotic is approved for clinical use.
Its in-vivo selection differential against the most common resistance
mutation is $s = 0.8\,\text{generation}^{-1}$ (susceptible cells die
during treatment, resistant cells grow). The mutation rate at the
relevant nucleotide is $\mu = 10^{-9}$ per generation. The treated
patient population is $10^7$ per year and each patient harbours a
$10^9$-cell bacterial population at the start of treatment. Estimate
the time, in years from approval, by which the resistant frequency in
the treated population exceeds 1\%.

\smallskip
\textit{Estimate before reading on.}

\smallskip
The expected number of pre-existing resistant cells in a $10^9$-cell
population, given mutation rate $\mu = 10^{-9}$, is one. Treatment of
$10^7$ patients per year means $10^7$ independent fixation events
per year, of which almost all start with one pre-existing resistant
mutant per patient. With $s = 0.8\,\text{generation}^{-1}$ and a
bacterial generation time of 30 minutes, $s = 0.8 \cdot 48 =
38\,\text{day}^{-1}$ at the patient level; in a 7-day course,
the resistant allele reaches fixation within each treated patient
before the course completes. The population-level frequency is
governed by the proportion of patients who acquire resistance under
treatment, which depends on combination-drug versus monotherapy,
treatment adherence, and transmission of resistant strains between
patients. Empirically, monotherapy with a single-step-resistance drug
reaches 1\% population-level resistance in 1--3 years, and 50\%
within a decade, on the historical record of the first-generation
$\beta$-lactams, the macrolides, and the fluoroquinolones.

\smallskip
\textit{Postmortem.} The estimate misses by a factor of 3--10 in
either direction for any single drug because four factors are not
in the simple selection-coefficient model: (1) horizontal gene
transfer can deliver pre-existing resistance plasmids from the
microbiota, dropping the effective $f_0$; (2) combination therapy
multiplies the required mutational supply, raising the effective
$f_0^{-1}$ by orders of magnitude; (3) clinical use beyond approved
indications (agricultural, over-the-counter in some jurisdictions)
accelerates selection in non-clinical reservoirs; (4) the cost of
the resistance mutation may be negligible after a single
compensatory mutation, so the steady-state frequency once
selection is removed is not $f_0$ but something larger. A working
heuristic: every new monotherapy antibiotic reaches clinically
significant resistance within 5--10 years; combination regimens
with two independent resistance pathways stretch the figure to
15--30 years.
\end{estimation}

\section{Conservation biology engineering-first}\pathtag{standard}

A small population faces three classes of risk \cite[ch.~3]{text:v6c08-caughley-conservation}:
\engterm{demographic stochasticity} (random variation in birth and
death events when $N$ is small enough that individual events matter),
\engterm{environmental stochasticity} (random variation in the
parameters $r$, $K$, $\beta$, $\gamma$ from year to year), and
\engterm{genetic deterioration} (inbreeding depression and loss of
adaptive variation when the effective population size is too small).
A conservation programme is an engineering intervention designed to
move a population away from one or more of these risks. The
discipline borrows the language of reliability engineering:
\engterm{minimum viable population} for the smallest $N$ at which
the extinction probability over a specified horizon falls below an
acceptable threshold (commonly 1\% over 100 years); \engterm{population
viability analysis} for the stochastic simulation that estimates
extinction probability under specified scenarios.

\engterm{Demographic stochasticity}. For a population with per-capita
birth rate $b$ and death rate $d$, the variance in $\Delta N$ over
one generation is approximately $(b + d) N$. The standard deviation
of the per-capita rate $\Delta N / N$ is therefore $\sqrt{(b + d)/N}$,
which scales as $N^{-1/2}$. For $N = 100$ and $b + d = 1\,
\text{yr}^{-1}$, the standard deviation of the per-capita growth
rate is 0.1 yr$^{-1}$, comparable to the per-capita growth rate
itself for a stable population. The implication is that small
populations have growth rates dominated by noise rather than by
mean trend; even a population with positive expected growth can go
extinct by demographic accident.

\engterm{Environmental stochasticity}. The variance in $r$ across
years has a different scaling. If $r$ varies year to year with
variance $\sigma_r^2$, the long-term geometric growth rate of the
population is approximately $\bar r - \sigma_r^2/2$, where $\bar r$ is
the arithmetic mean. A population with positive mean growth rate and
high environmental variance can have negative long-term growth: the
arithmetic and geometric means diverge for variable populations. The
engineering implication is that the relevant quantity for long-term
viability is the geometric mean of annual growth rates, not the
arithmetic mean, and reducing year-to-year variance (by buffering
extreme conditions, by maintaining metapopulation structure) can be
as effective as raising the mean.

\engterm{Genetic deterioration}. A population with effective size $N_e$
loses heterozygosity at the rate $1/(2 N_e)$ per generation by drift,
and accumulates inbreeding at the same rate. For $N_e = 50$, the
loss is 1\% per generation; the rule of thumb that $N_e \geq 50$
prevents short-term inbreeding depression and $N_e \geq 500$
prevents long-term loss of adaptive variation is one of the few
operational thresholds the field has \cite[ch.~7]{text:v6c08-caughley-conservation}.
The effective population size $N_e$ is typically much smaller than
the census size $N$, with the ratio $N_e/N$ commonly 0.1 to 0.3 for
vertebrate populations because of variance in reproductive success,
fluctuating population size, and unequal sex ratio.

\engterm{Metapopulations and reserve design}. A metapopulation is a
collection of local populations connected by migration. Levins's
1969 model \cite[pp.~237--240]{paper:v6c08-levins-1969-metapopulation}
treats the fraction of habitable patches occupied at time $t$:
\[
\frac{dp}{dt} = c p (1 - p) - e p,
\]
with $c$ the colonisation rate per occupied patch into empty
patches, $e$ the local extinction rate per occupied patch. The
equilibrium occupancy is $p^* = 1 - e/c$, positive when $c > e$. The
metapopulation persists at the regional scale even if every local
population is going extinct, provided the colonisation rate exceeds
the extinction rate. Reserve design for endangered species uses
this framing: a network of small patches with good connectivity may
support a metapopulation that single large patches alone could not.

\section{Failure: ecosystem collapse and the pandemic-prediction record}\pathtag{core}

Models of populations, ecosystems, and epidemics fail in
characteristic ways. The chapter's failure section names four.

\engterm{Logistic-model failure: overshoot and collapse}. The
classical logistic equation predicts a smooth approach to carrying
capacity from any positive initial condition. Real populations on
finite resources frequently overshoot the steady carrying capacity,
degrade the resource, and crash. The Tasmanian sheep population, the
St.\ Matthew Island reindeer, the Easter Island human population
\cite[chs.~2--3]{text:v6c08-diamond-collapse}, and the Atlantic cod
fishery of the Grand Banks \cite[ch.~9]{text:v6c08-kurlansky-cod} are the
named cases the chapter draws on. The mechanism is that $K$ is not
fixed: the population's activity degrades the resource, and a
population that overshoots the long-run sustainable $K$ degrades the
resource faster than it regenerates. The engineering lesson is to
identify the resource time scale, to monitor it directly rather than
inferring it from population size, and to set harvest or growth
limits below the static estimate of $K$ to allow for delay.

The Easter Island reading is contested. Diamond's 2005 narrative
makes Rapa Nui a parable of self-inflicted ecological suicide: a
human population that arrived around 800 CE deforested the island
for moai-transport rollers and agricultural clearance, lost soil
fertility and freshwater catchment, and crashed before European
contact. Hunt and Lipo's 2011 revisionist account
\cite[chs.~4--6]{text:v6c08-hunt-lipo-2011}, building on
radiocarbon-dating evidence published in \emph{Science} in 2006
\cite[pp.~1603--1606]{paper:v6c08-hunt-lipo-rapanui-2006}, redates
arrival to circa 1200 CE, attributes most deforestation to the
introduced Polynesian rat consuming palm seeds, and argues the
population was relatively stable at 3000--4000 individuals until
European contact in 1722, after which European-introduced disease
and slave-raiding drove the collapse documented in nineteenth-century
records. The two readings carry different engineering lessons: the
Diamond reading reinforces the overshoot-and-collapse model with $K$
endogenous to population activity; the Hunt-Lipo reading shifts the
collapse mechanism to introduced disease and external predation,
leaving the pre-contact dynamics closer to an Allee-effect or
metapopulation case. Both readings are taught here as live; the
reader who claims certainty about Easter Island has not read the
contested literature.

The Atlantic cod collapse is less contested. Hutchings and Myers'
1994 reconstruction \cite[pp.~2126--2146]{paper:v6c08-hutchings-cod-1996}
documents the harvest history of the northwest Atlantic cod stocks
from 1500 CE to the 1992 moratorium. Catches sustained at
$0.2 \times 10^6\,\text{t}/\text{yr}$ through the sixteenth and
seventeenth centuries; rose with steam-trawler technology to
$1$--$1.5 \times 10^6\,\text{t}/\text{yr}$ in the early twentieth
century; spiked under foreign factory-trawler fleets to $2$--$2.5
\times 10^6\,\text{t}/\text{yr}$ in the 1960s; and collapsed
through the 1980s as spawning-stock biomass fell below $1\%$ of
historical levels. Myers, Hutchings, and Barrowman's 1997 follow-up
\cite[pp.~91--106]{paper:v6c08-myers-cod-1994} demonstrated that the
stock-recruitment curve had shifted: the relationship between
spawning-stock biomass and recruitment of young fish, fitted to
post-1962 data, gave a recruitment-replacement equilibrium far below
the harvest rate set by the Canadian Department of Fisheries and
Oceans throughout the 1980s. The collapse is the textbook case of
MSY-target management combined with optimistic $K$ estimation;
the resource time scale (cod reach reproductive age at 5--6 years)
was longer than the data-revision cycle (annual catches, biennial
stock assessments), and the assessment lag is a load-bearing
component of the failure. Three decades after the moratorium, the
stock has partially recovered in the southern Gulf of St.\ Lawrence
and remained at $<5\%$ of historical biomass on the northern Grand
Banks. Jackson et al.\ 2001 in \emph{Science}
\cite[pp.~629--637]{paper:v6c08-jackson-marine-2001} embedded the
cod collapse in the broader baseline-shift problem for coastal
ecosystems: every generation of fisheries managers since the
nineteenth century has treated the early-career baseline as the
historical norm, with the cumulative result that the eighteenth-
century carrying capacity has dropped from view.

The passenger pigeon case (the most extreme Allee-driven extinction
in the modern record) belongs to the same family. The species's
total population in North America was estimated at 3--5 billion in
1850; the last wild bird was shot in 1900; the last individual
(Martha) died at the Cincinnati Zoo in 1914. The proximate cause was
commercial market hunting enabled by the telegraph (which
synchronised hunting parties to migration flocks) and the railroad
(which moved birds to urban markets). The mechanistic story
combines harvest mortality with an Allee effect: the species's
reproductive biology required colonial nesting at densities above
some threshold (mate-finding, predator satiation), and as densities
fell below that threshold the per-capita reproductive rate
collapsed. The model that fits the data is logistic-with-harvest
plus a strong Allee threshold; the population fell into the
sub-threshold basin from which no harvest-cessation policy could
have recovered it, even when policy belatedly arrived in 1900.

\engterm{Lotka-Volterra failure: neutral stability is not real}. The
original predator-prey model produces neutral oscillations: every
initial condition gives a different limit cycle, with no damping.
Empirical predator-prey systems do not show neutral stability. They
show damped oscillations toward a stable equilibrium (often), stable
limit cycles (sometimes), or extinction (occasionally). The model
fails because its assumptions (linear prey response, linear predator
response, no carrying capacity for prey) do not hold. The
Rosenzweig-MacArthur model with logistic prey and Holling type II
predator functional response repairs the failure; the engineering
lesson is that a model whose structural prediction (neutral
stability) is not observed must be modified rather than retained.

\engterm{SIR failure modes: heterogeneity}. The basic SIR model
assumes homogeneous mixing. Real populations are heterogeneous in
contact rates: some individuals interact with many others (the
``superspreaders'' identified during the SARS-CoV-1 outbreak of
2003 and again during SARS-CoV-2 in 2020--2022), while most
interact with few. Heterogeneity has two effects: the effective
$R_0$ is dominated by the high-contact tail of the distribution, so
naive average-based estimates understate $R_0$; the herd-immunity
threshold is lower than the homogeneous-mixing calculation predicts,
because the high-contact individuals are the first to be infected
and removed from the susceptible pool
\cite[Tab.~2]{paper:v6c08-lloyd-smith-2005-superspreaders}. The 2020--2021
public discourse used homogeneous SIR predictions without
adjustment and consistently overestimated the herd-immunity
fraction.

\input{volumes/06-life/ch08-population-ecology-epidemiology/figures/fig-pandemic-timeline.tex}

\input{volumes/06-life/ch08-population-ecology-epidemiology/figures/fig-covid-incidence.tex}

\engterm{COVID-19 modelling: the public record}. The early SARS-CoV-2
models of 2020 produced wildly divergent forecasts. The Imperial
College modelling group's March 2020 report
\cite[pp.~5--7]{paper:v6c08-ferguson-imperial-r9-2020} projected, under
no-intervention scenarios, between 500{,}000 and 2{,}200{,}000
deaths in the United Kingdom, depending on whether infection
fatality rate was at the low or high end of available estimates.
The IHME model used phenomenological curve-fitting rather than
mechanistic SIR and produced peak-deaths predictions that varied by
factors of two to ten across successive weekly updates. The
mechanistic models in turn varied by a factor of three or more
across institutions and across weeks. The retrospective on the
modelling effort is mixed: peak timing was forecast within a few
weeks in most cases; final size and total deaths were forecast
poorly, often by factors of two or three; the relative effects of
non-pharmaceutical interventions (school closures, masking,
distance limits) were forecast with very wide error bars
\cite{paper:v6c08-adam-2020-models}. The engineering lessons are
specific: at the early-outbreak phase, the prediction is dominated
by the parameter uncertainty (especially in the case fatality rate
and the contact heterogeneity), not by the model structure; reports
that compress this uncertainty into a single number mislead; the
appropriate output of an epidemic model in a fast-moving outbreak is
an envelope of scenarios, not a forecast.

Three specific failures recur in the post-2020 modelling
retrospectives. The first is testing-capacity confounding. Costello,
Pollock, Roderick, and Sridhar's 2020 \emph{BMJ} review
\cite[pp.~m1567-1--m1567-3]{paper:v6c08-costello-testing-2020}
documented that early case-count time series in the UK, Spain,
Italy, and the US tracked testing capacity at least as much as
underlying incidence: when testing capacity doubled in a week, the
case count appeared to double whether or not transmission accelerated.
A model that fits $\beta$ to such a series infers an inflated
transmission rate and over-predicts the eventual peak. The
engineering correction is to fit to deaths (lagged by 2--3 weeks),
to hospitalisations, or to wastewater-surveillance signals; or to
include a time-varying ascertainment fraction as an inferred
parameter, with a prior tied to test-capacity records. The
correction was applied unevenly across 2020--2021 forecasting
shops; outputs that did not apply it (most published forecasts in
March--May 2020) carried errors of the order of the testing-capacity
shifts, not the actual transmission dynamics.

The second failure is the homogeneous-mixing assumption applied to
herd-immunity-threshold calculations. The Lloyd-Smith et al.\ 2005
superspreader paper had been in the literature for fifteen years
before SARS-CoV-2 arrived; early-pandemic public communications in
the United Kingdom, Sweden, and the United States nonetheless quoted
homogeneous-mixing thresholds (60--70\%) without the heterogeneity
correction that would have brought the effective threshold to
40--50\% under reasonable assumptions about contact-rate variance.
The communication failure was not in the original modelling papers,
which used heterogeneous structures appropriate to the question;
the failure was in the translation to policy briefings, which
flattened the structural assumption.

The third failure is the late incorporation of immune waning and
variant emergence. Models fitted in mid-2020 implicitly assumed a
single pathogen with permanent immunity on recovery. The first
revisions for variant emergence appeared in late 2020 (Alpha) and
were routine by mid-2021 (Delta); the framework for immune waning
was incorporated systematically only after winter 2021--2022 (when
breakthrough infections after Pfizer-BNT162b2 became a measurable
phenomenon). Eddington's 2020 \emph{Lancet Public Health}
commentary \cite[pp.~e379--e380]{paper:v6c08-eddington-lancet-2020}
made the case that the next-pandemic modelling framework should
budget for variant emergence and immune waning from the start, with
default parameters drawn from the SARS-CoV-2 record rather than
from the assumption of a single fixed pathogen.

\begin{principle}[Models with structural assumptions outside their domain mislead]
A model's predictive power depends on its structural assumptions
holding within the domain where the prediction is wanted. Logistic
models with fixed $K$, Lotka-Volterra models with no carrying
capacity, and SIR models with homogeneous mixing all break when
their assumptions are violated. The engineering response is not to
trust models less but to identify in advance which assumption is
load-bearing, to monitor the assumption empirically, and to retire
the model when its assumption fails.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Fit an SIR model to a real epidemic dataset]
The reader fits an SIR or SEIR model to a published dataset for a
real outbreak, reports parameter estimates with uncertainties, and
documents the model's predictive limits.

\textbf{Track}. Simulation.
\textbf{Hazard class}: low (no pathogen handling; published data
only).

\textbf{Dataset options}.

\begin{itemize}[leftmargin=*]
\item Boarding-school influenza, England 1978 (a small closed
  population with daily case counts; the textbook teaching case).
\item Eyam plague, England 1665--1666 (a small village with detailed
  records; SIR's original test case in retrospect).
\item Ebola outbreak in West Africa 2014--2015 (WHO weekly case
  counts; published mechanistic-modelling papers as benchmark).
\item SARS-CoV-2 first-wave data for a single country or region,
  through to first-wave peak (Johns Hopkins University CSSE archive
  or country-specific official source).
\end{itemize}

\textbf{Procedure}.

\begin{enumerate}[leftmargin=*]
\item Identify the dataset and the time window of analysis. State
  the data source, the time resolution, and the case definition.
\item Choose a model structure (SIR, SEIR, age-stratified, with or
  without waning immunity). Justify the choice from the dataset's
  characteristics and the disease's biology.
\item Estimate $R_0$ from the early-phase exponential growth rate.
  Estimate the infectious period $1/\gamma$ from independent clinical
  data or from the published meta-analyses.
\item Fit the model to the case-incidence time series by least
  squares or by maximum likelihood. Report parameter point estimates
  and 95\% confidence intervals.
\item Compare predicted and observed peak time, peak height, and
  final size. Report the residuals as a time series and assess
  whether they are consistent with the model's stochastic
  assumptions.
\item Identify two structural assumptions of the model that are
  likely violated by the dataset, and discuss how the violations
  would affect the parameter estimates.
\end{enumerate}

\textbf{Deliverable}. A 10- to 15-page report with one parameter
table, two figures (case-incidence fit and residuals), and a
postmortem section. Include the code in an appendix or as a
publicly available script.

\textbf{Time}. 10--20 hours of work, depending on dataset complexity.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A bacterial culture grows from $10^3$ cells/mL to $10^6$ cells/mL in
6 hours. Compute the per-capita growth rate $r$ and the doubling
time.

\paragraph{Solution sketch.} Exponential growth gives
$N(t) = N_0 e^{rt}$, so
$r = \ln(N/N_0)/t = \ln(10^3)/6\,\text{h} = 6.91/6 = 1.15\,
\text{h}^{-1}$. The doubling time is $T_d = \ln 2 / r =
0.693/1.15 = 0.60\,\text{h}$, about 36 minutes, consistent with
\emph{E.\ coli} in log phase.
\end{exercise}

\begin{exercise}
A population with $r = 0.05\,\text{yr}^{-1}$ and $K = 1000$ starts at
$N_0 = 50$. Use the logistic solution to compute $N(t)$ at $t = 10$,
20, 50, and 100 years.

\paragraph{Solution sketch.} The logistic solution is
$N(t) = K / (1 + A e^{-rt})$ with $A = (K - N_0)/N_0 = 950/50 = 19$.
Evaluating: $N(10) = 1000/(1 + 19 e^{-0.5}) = 1000/(1 + 11.52) =
80$; $N(20) = 1000/(1 + 19 e^{-1.0}) = 1000/(1 + 6.99) = 125$;
$N(50) = 1000/(1 + 19 e^{-2.5}) = 1000/(1 + 1.56) = 391$; $N(100) =
1000/(1 + 19 e^{-5}) = 1000/(1 + 0.128) = 887$. The population
crosses $K/2 = 500$ near $t = 59\,\text{yr}$ and approaches $K$
asymptotically.
\end{exercise}

\begin{exercise}
For the Lotka-Volterra equations with $\alpha = 1.0\,\text{yr}^{-1}$,
$\beta = 0.01\,(\text{predator}\cdot\text{yr})^{-1}$, $\delta =
0.005\,(\text{prey}\cdot\text{yr})^{-1}$, $\gamma = 0.5\,\text{yr}^{-1}$,
compute the equilibrium prey and predator densities and the period of
small oscillations.

\paragraph{Solution sketch.} $N^* = \gamma/\delta = 0.5/0.005 =
100$ prey; $P^* = \alpha/\beta = 1.0/0.01 = 100$ predators. The
small-oscillation period is
$T = 2\pi/\sqrt{\alpha\gamma} = 2\pi/\sqrt{0.5} = 8.89\,\text{yr}$,
in the right order for predator-prey cycles in long-lived
vertebrates (lynx-hare at $\approx 10\,\text{yr}$, snowshoe-hare at
similar period).
\end{exercise}

\begin{exercise}
A disease has infectious period 5 days and $R_0 = 2.5$. Compute
$\beta$ and $\gamma$. Compute the herd-immunity threshold. Compute
the final fraction infected in an unmitigated outbreak.

\paragraph{Solution sketch.} $\gamma = 1/D = 1/5 =
0.20\,\text{d}^{-1}$. From $R_0 = \beta/\gamma = 2.5$,
$\beta = 0.50\,\text{d}^{-1}$. Herd-immunity threshold:
$p_c = 1 - 1/R_0 = 1 - 0.4 = 0.60$ or 60\%. Final fraction
infected solves $1 - s_\infty = 1 - e^{-2.5(1 - s_\infty)}$;
iterating from $s_\infty^{(0)} = 0.1$ gives $s_\infty \approx
0.107$, so the final attack rate is $\approx 89.3\%$. The mismatch
(60\% threshold versus 89\% final size) is the standard
overshoot phenomenon: the epidemic continues past the threshold
while infectious individuals exist, infecting additional
susceptibles before $I$ falls to zero.
\end{exercise}

\begin{exercise}
The early-phase exponential growth rate of an outbreak is
$\lambda = 0.20\,\text{day}^{-1}$. The infectious period is 4 days.
Estimate $R_0$.

\paragraph{Solution sketch.} $\gamma = 1/4 = 0.25\,\text{d}^{-1}$.
For an SIR model, $\lambda = \beta - \gamma$, so $\beta = \lambda
+ \gamma = 0.45\,\text{d}^{-1}$ and $R_0 = \beta/\gamma = 1.80$.
The shortcut form $R_0 \approx 1 + \lambda/\gamma = 1 + 0.80 = 1.80$
gives the same answer. An SEIR model with a 3-day latent period
would yield a slightly higher $R_0$ (around 2.0--2.2) for the same
observed $\lambda$, because the latent compartment delays
exponential growth.
\end{exercise}

\begin{exercise}
For a Leslie matrix with three age classes and fertilities
$F_1 = 0$, $F_2 = 2$, $F_3 = 4$ and survivals $P_1 = 0.7$, $P_2 =
0.5$, compute the dominant eigenvalue and the stable age
distribution.

\paragraph{Solution sketch.} The Leslie matrix is
\[
L = \begin{pmatrix} 0 & 2 & 4 \\ 0.7 & 0 & 0 \\ 0 & 0.5 & 0
\end{pmatrix}.
\]
The characteristic polynomial is
$-\lambda^3 + 2 \cdot 0.7\,\lambda + 4 \cdot 0.7 \cdot 0.5 = 0$,
i.e.\ $\lambda^3 - 1.4\lambda - 1.4 = 0$. Iterating Newton's method
from $\lambda = 1.5$ gives $\lambda_1 \approx 1.452$, the
asymptotic growth rate per time step. The right eigenvector
satisfies $L\mathbf{v} = \lambda \mathbf{v}$. From row 2,
$v_2 = 0.7 v_1 / \lambda$; from row 3, $v_3 = 0.5 v_2 / \lambda =
0.35 v_1 / \lambda^2$. Normalising so $v_1 + v_2 + v_3 = 1$ with
$\lambda = 1.452$: $v_2 = 0.482 v_1$, $v_3 = 0.166 v_1$,
$v_1 + v_2 + v_3 = 1.648 v_1 = 1$, so $\mathbf{v} \approx (0.607,
0.293, 0.101)$. The population grows at $\sim 45\%$ per time step
in its stable age distribution.
\end{exercise}

\begin{exercise}
A bacterial population of $10^9$ cells contains a single resistant
mutant ($f_0 = 10^{-9}$). The selection differential in the presence
of drug is $r_R + d_S = 0.8\,\text{h}^{-1}$. Compute the time for the
resistant fraction to rise from $10^{-9}$ to $10^{-1}$.

\paragraph{Solution sketch.} The logistic equation in the resistant
fraction integrates to $f(t)/(1 - f(t)) = (f_0/(1 - f_0)) e^{st}$
with $s = r_R + d_S = 0.8\,\text{h}^{-1}$. For $f_0 = 10^{-9}$ and
target $f = 10^{-1}$,
\[
t = \frac{1}{s} \ln \frac{f/(1-f)}{f_0/(1-f_0)} =
\frac{1}{0.8} \ln \frac{0.111}{10^{-9}} \approx \frac{18.5}{0.8}
\approx 23\,\text{h}.
\]
Under a day for selection to dominate the population from a single
pre-existing mutant. Combination therapy with two independent
resistances pushes $f_0$ to $10^{-18}$ and the time scale to $\sim
44\,\text{h}$ for the same target, still under two days; the
multiplicative benefit comes from the requirement for two
independent fixation events.
\end{exercise}

\begin{exercise}
A metapopulation has colonisation rate $c = 0.3\,\text{yr}^{-1}$ per
occupied patch and extinction rate $e = 0.1\,\text{yr}^{-1}$ per
occupied patch. Compute the equilibrium occupancy. What value of
$e/c$ produces extinction?

\paragraph{Solution sketch.} From $dp/dt = cp(1-p) - ep = 0$ at
$p = p^*$, the equilibrium is $p^* = 1 - e/c = 1 - 0.1/0.3 = 2/3$.
The trivial equilibrium $p^* = 0$ is unstable when $c > e$. The
metapopulation goes extinct (the only stable equilibrium is
$p^* = 0$) when $e \geq c$, i.e.\ when the local extinction rate
exceeds the colonisation rate. The implication: habitat
fragmentation that reduces $c$ (longer between-patch distances,
weaker dispersal) without changing $e$ flips the system from
persistence to extinction at $c = e$.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the final-size relation $1 - s_\infty = 1 - e^{-R_0(1 -
s_\infty)}$ for the SIR model by integrating the equations from
$t = 0$ (with $S = N$, $I = I_0$) to $t = \infty$ (with $I = 0$).

\paragraph{Solution sketch.} Divide $dS/dt$ by $dR/dt$ to eliminate
time: $dS/dR = -\beta S/(\gamma N) = -R_0 S/N$, since $\beta/\gamma
= R_0$. Integrate to get $\ln(S/N) = -R_0 (R - R_0^{\mathrm{init}})/N$,
or with the initial condition $S(0) = N - I_0 \approx N$ and
$R(0) = 0$: $S(t)/N = \exp(-R_0 R(t)/N)$. As $t \to \infty$, $I \to
0$ and $R(\infty)/N = 1 - s_\infty$ (since $S(\infty)/N = s_\infty$
and $S + R = N$ at infinity). Substituting, $s_\infty =
\exp(-R_0(1 - s_\infty))$, which rearranges to the stated form.
For $R_0 = 1$ the only solution is $s_\infty = 1$ (no outbreak);
for $R_0 > 1$ a non-trivial root exists.
\end{exercise}

\begin{exercise}
Derive the conservation law $S + I + R = N$ from the SIR equations
and show it is consistent with the equations' structure.

\paragraph{Solution sketch.} Add the three equations:
$d(S+I+R)/dt = -\beta SI/N + \beta SI/N - \gamma I + \gamma I = 0$.
So $S + I + R$ is conserved; if the initial conditions sum to $N$,
they sum to $N$ at all times. The conservation reflects the closed
population: no births, no deaths, no immigration, no emigration.
Extensions: SEIR adds an $E$ compartment with rate $\sigma E$ in
$dI/dt$ and $-\sigma E$ in $dE/dt$, preserving conservation. SEIRD
adds a $D$ (dead) compartment with $\mu I$ in $dD/dt$ and
$-(\gamma + \mu) I$ in $dI/dt$; total $S + E + I + R + D = N$ is
conserved, but $S + E + I + R$ is not (the closed-system check
becomes: the population includes the dead).
\end{exercise}

\begin{exercise}
Derive the linear-stability analysis of the Lotka-Volterra
equilibrium $(N^*, P^*)$. Show that the eigenvalues are imaginary,
giving neutral stability, and compute the small-oscillation
frequency.

\paragraph{Solution sketch.} The Jacobian of
$(f_1, f_2) = (\alpha N - \beta NP, \delta NP - \gamma P)$ is
\[
J = \begin{pmatrix} \alpha - \beta P & -\beta N \\
\delta P & \delta N - \gamma \end{pmatrix}.
\]
At $(N^*, P^*) = (\gamma/\delta, \alpha/\beta)$, the diagonal
entries vanish:
\[
J^* = \begin{pmatrix} 0 & -\beta N^* \\
\delta P^* & 0 \end{pmatrix} =
\begin{pmatrix} 0 & -\beta\gamma/\delta \\
\delta\alpha/\beta & 0 \end{pmatrix}.
\]
The characteristic polynomial is $\lambda^2 + \alpha\gamma = 0$, so
$\lambda = \pm i\sqrt{\alpha\gamma}$. Imaginary eigenvalues mean
neutral stability (a centre, not a focus); the small-oscillation
angular frequency is $\omega = \sqrt{\alpha\gamma}$, and the period
is $T = 2\pi/\sqrt{\alpha\gamma}$, matching the period quoted in
§8.2.
\end{exercise}

\begin{exercise}
Derive the Rosenzweig-MacArthur model from the Lotka-Volterra model
by replacing the linear prey growth with logistic growth and the
linear predator response with Holling type II. State the conditions
for a stable equilibrium and for a stable limit cycle (the
``paradox of enrichment'' condition).

\paragraph{Solution sketch.} Replace prey term $\alpha N \to rN(1 -
N/K_N)$ and predation term $\beta NP \to \beta NP/(N + h)$. The
predator equation uses the same saturating form
$\delta NP/(N + h) - \gamma P$. The non-trivial equilibrium has
$N^* = \gamma h / (\delta - \gamma)$ and $P^*$ obtained from the
prey equation at equilibrium. Linear-stability analysis at
$(N^*, P^*)$ produces a $2 \times 2$ Jacobian whose trace depends on
$K_N$. As $K_N$ increases past a critical value
$K_N^{\mathrm{crit}} = h(\delta + \gamma)/(\delta - \gamma)$, the
trace changes sign: below, the trace is negative and the equilibrium
is a stable focus; above, the trace is positive and the equilibrium
is unstable, with a stable limit cycle emerging through a
supercritical Hopf bifurcation. The paradox-of-enrichment condition
is $K_N > K_N^{\mathrm{crit}}$. The implication: improving the
prey's environment can destabilise the predator population, an
unwelcome surprise for the conservationist who raises $K_N$ to
help the prey but inadvertently increases predator oscillation
amplitude and the probability of extinction-by-noise.
\end{exercise}

\begin{exercise}
Derive the long-term geometric growth rate $\bar r - \sigma_r^2/2$
for a population with year-to-year variance $\sigma_r^2$ in its
per-capita growth rate. Use the log-normal approximation.

\paragraph{Solution sketch.} If $r_t$ is the per-capita growth rate
in year $t$, the population after $n$ years is
$N_n = N_0 \prod_{t=1}^{n} e^{r_t} = N_0 \exp\left(\sum_t
r_t\right)$. The long-term per-step log-growth rate is
$\lim_{n \to \infty} (1/n) \sum_t r_t = \E[r]$, so the geometric
growth rate of $N$ itself is $e^{\E[r]}$. But the \emph{arithmetic}
expected growth rate is $\E[e^{r_t}]$. If $r_t$ is normally
distributed with mean $\bar r$ and variance $\sigma_r^2$, then by
the moment-generating function for the normal,
$\E[e^{r_t}] = e^{\bar r + \sigma_r^2/2}$, so the arithmetic-mean
per-step growth factor is $e^{\bar r + \sigma_r^2/2}$ but the
geometric (long-term) factor is $e^{\bar r}$. In log terms, the
long-term mean per-step log-growth is $\bar r$, while the
arithmetic mean over realisations is $\bar r + \sigma_r^2/2$. The
difference $\sigma_r^2/2$ is the variance correction; a population
with $\bar r > 0$ but variance high enough that $\bar r <
\sigma_r^2/2$ has positive arithmetic growth but negative geometric
growth and goes extinct in the long run despite positive expected
year-on-year change.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the time for a measles outbreak to infect the
herd-immunity fraction of a fully susceptible population of
$10^6$ people, given $R_0 = 15$ and infectious period 8 days.

\paragraph{Solution sketch.} Herd-immunity threshold
$p_c = 1 - 1/R_0 = 14/15 \approx 0.933$, so $\sim 9.3 \times 10^5$
infections. Early-phase growth rate $\lambda = (R_0 - 1)/D =
14/8 = 1.75\,\text{d}^{-1}$, doubling time $\ln 2/\lambda =
0.40\,\text{d}$ (about 10 hours). To reach the threshold from
$I_0 = 1$, infected count must reach order $10^5$ before
saturation begins to bite; in pure exponential growth, time is
$\ln(10^5)/\lambda \approx 11.5/1.75 = 6.6\,\text{d}$. Including
the slowdown from depleting susceptibles, the peak of the
unmitigated epidemic arrives near day 18--22 and the herd-immunity
fraction is crossed near day 25--30. The estimate aligns with the
historical record of measles in island populations before the
vaccine era (Faroe Islands 1846 outbreak, Fiji 1875 outbreak),
where unmitigated outbreaks ran their course in 4--8 weeks.
\end{exercise}

\begin{exercise}
Estimate the carrying capacity (in cells per millilitre) of a
laboratory yeast culture in 100\,mL of glucose-rich medium, given
that each yeast cell consumes about $10^{-12}$\,g of glucose per
generation and the medium contains 50\,g/L of glucose.
\paragraph{Solution sketch.} Total glucose available is
$0.1\,\text{L} \cdot 50\,\text{g/L} = 5\,\text{g}$. If each cell
consumes $10^{-12}\,\text{g}$ per generation and cells must sustain
metabolism through $\sim 20$ generations before reaching stationary
phase, total consumption per surviving cell is $\sim 2 \times
10^{-11}\,\text{g}$. The maximum number of cells the medium
supports is $5 / (2 \times 10^{-11}) = 2.5 \times 10^{11}$ cells
in 100\,mL, or $K \approx 2.5 \times 10^9\,\text{cells/mL}$,
consistent with reported stationary-phase yeast densities of
$1\text{--}5 \times 10^9\,\text{cells/mL}$ in YPD medium. The
estimate is sensitive to the maintenance-respiration overhead;
treating cells as one-shot consumers (a single $10^{-12}$ g per
cell) gives $K \approx 5 \times 10^{10}\,\text{cells/mL}$, a factor
of 20 over-estimate.
\end{exercise}

\begin{exercise}
Estimate the demographic-stochasticity-induced standard deviation
in the annual growth rate of a population of $N = 200$ individuals
with $b + d = 0.5\,\text{yr}^{-1}$.

\paragraph{Solution sketch.} For a birth-death process, the
variance in the change $\Delta N$ over one generation is
approximately $(b + d) N$, so the standard deviation of $\Delta
N/N$ is $\sqrt{(b + d)/N} = \sqrt{0.5/200} = \sqrt{0.0025} =
0.050\,\text{yr}^{-1}$, or 5\% per year. The implication: a
population of 200 with an expected growth rate of, say, $r = 0.03\,
\text{yr}^{-1}$ has demographic stochasticity (5\%) larger than
the mean trend (3\%), so the observed time series will be
dominated by year-to-year random walk rather than by exponential
recovery. A doubling to $N = 800$ halves the stochasticity to
2.5\%, bringing the mean trend back to the dominant signal.
\end{exercise}

\begin{exercise}
Estimate the effective population size $N_e$ of a vertebrate
population with census size $N = 1000$, sex ratio 1:1, and a
coefficient of variation in reproductive success of 1.5. Use the
relation $N_e \approx 4 N_f N_m / (N_f + N_m)$ corrected by the
variance term.

\paragraph{Solution sketch.} With $N_f = N_m = 500$, the sex-ratio
formula gives $N_e^{\mathrm{sex}} \approx 4 \cdot 500 \cdot 500/
(500 + 500) = 1000$, equal to the census; balanced sex ratios do
not lose effective size. The variance correction is
$N_e \approx N_e^{\mathrm{sex}}/(1 + \mathrm{CV}^2)$ where CV is the
coefficient of variation in reproductive success. With $\mathrm{CV}
= 1.5$, $1 + \mathrm{CV}^2 = 1 + 2.25 = 3.25$, so
$N_e \approx 1000/3.25 \approx 308$. The ratio $N_e/N = 0.31$ is
in the typical vertebrate range (0.1--0.5). For conservation
purposes, the relevant census size for genetic management is the
effective size, not the census; a wild population of 1000 with
this variance behaves genetically like a population of 308.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement the SEIR model in a numerical ODE solver. Run it for
$R_0 = 3$, $1/\sigma = 5$ days, $1/\gamma = 7$ days, $N = 10^6$,
$I_0 = 10$. Compute the peak time, peak prevalence, and final
attack rate.

\paragraph{Solution sketch.} Use \texttt{code/sir\_seir\_solve.py}
(included in this chapter's \texttt{code/} directory) or
\texttt{scipy.integrate.solve\_ivp} with the SEIR equations from
§8.3. With $\beta = R_0 \gamma = 3 \cdot (1/7) = 0.429\,
\text{d}^{-1}$ and $\sigma = 0.2\,\text{d}^{-1}$, the simulation
gives peak time $\approx 90\,\text{days}$ (compared to $\sim 65\,
\text{days}$ for the equivalent SIR), peak prevalence
$I_{\max}/N \approx 0.165$ (about $1.65 \times 10^5$ simultaneous
infectious cases), and final attack rate $1 - s_\infty \approx
0.94$ (the SIR final-size relation is unchanged by the latent
compartment, only the dynamics on the way are delayed). The latent
period lowers the peak by $\approx 3\%$ relative to SIR and shifts
peak time later by $\approx 25\,\text{days}$ at these parameters.
\end{exercise}

\begin{exercise}
Implement the stochastic version of the SIR model using the Gillespie
algorithm. Run 1000 realisations for $R_0 = 1.5$, $N = 100$, $I_0 =
1$. Report the fraction of realisations in which the outbreak
fades out before infecting more than 10 individuals.

\paragraph{Solution sketch.} The probability of extinction in the
branching-process approximation is $1/R_0$ per initial infectious
individual, so $\approx 0.667$ for $R_0 = 1.5$. The Gillespie
simulation tracks discrete events; an outbreak that escapes the
early stochastic phase grows logistically. For $R_0 = 1.5$, $N =
100$, $I_0 = 1$, the fraction of small-outbreak realisations
(final size $\leq 10$) is empirically $\approx 0.55\text{--}0.65$
across 1000 runs, with the remaining $\approx 0.40$ producing
final attack rates between 50\% and 75\% (governed by the SIR
final-size relation $1 - s_\infty \approx 0.58$ at $R_0 = 1.5$).
The small-$N$ correction means a deterministic SIR fit to a small
outbreak overstates the true epidemic risk; the Gillespie estimate
should be the default for $N < 1000$ and $R_0 < 2$.
\end{exercise}

\begin{exercise}
Simulate the Rosenzweig-MacArthur model. Sweep the prey carrying
capacity $K$ and identify the value at which the stable
equilibrium becomes unstable and gives way to a stable limit cycle
(the Hopf bifurcation). Verify the paradox of enrichment.

\paragraph{Solution sketch.} The reference parameter set
$r = 0.5\,\text{yr}^{-1}$, $\beta = 0.02\,\text{yr}^{-1}\cdot
\text{prey}^{-1}$, $h = 50\,\text{prey}$, $\delta = 0.5 \beta$,
$\gamma = 0.1\,\text{yr}^{-1}$ gives
$N^* = \gamma h/(\delta - \gamma) = 0.1 \cdot 50/(0.01 - 0.1)
\Rightarrow$ check sign; use parameters where $\delta > \gamma$,
e.g.\ $\delta = 0.2 \beta$ and $\gamma$ smaller. A working set:
$r = 1$, $\beta = 0.5$, $h = 0.5$, $\delta = 0.4$, $\gamma = 0.2$,
$N^* = 0.5$. Sweeping $K$ from 0.5 to 4, the equilibrium is stable
for $K \lesssim 1.5$ and the limit cycle takes over for $K \gtrsim
1.5$. Use \texttt{code/lotka\_volterra.py} for the SciPy
implementation. The verification: $K_N^{\mathrm{crit}} \approx
h(\delta + \gamma)/(\delta - \gamma) = 0.5 \cdot 0.6/0.2 = 1.5$,
matching the simulated bifurcation point.
\end{exercise}

\begin{exercise}
Build a stochastic individual-based model of antibiotic resistance.
Start with $10^9$ susceptible cells, mutation rate $10^{-9}$ per
generation, selection differential $1\,\text{h}^{-1}$. Run the
simulation under constant-drug, pulsed-drug, and combination-drug
regimens and compare time to resistance fixation.

\paragraph{Solution sketch.} See \texttt{code/resistance\_evolution.py}.
Constant-drug regimen: fixation in $\sim 20\,\text{h}$ (the
single-mutation calculation from §8.5). Pulsed-drug regimen (drug
on/off cycles of 12\,h): fixation is delayed by the off-drug
intervals during which susceptible cells outgrow resistant cells
(fitness-cost-induced), reaching fixation in 36--60\,h depending
on the fitness cost and pulse duty cycle. Combination drug with two
independent resistance mutations (effective $f_0 = 10^{-18}$):
fixation in $\sim 42\,\text{h}$ for the same per-mutation
selection, the modest improvement reflecting the doubling of the
required $\ln(1/f_0)$ rather than a true multiplicative effect on
time. The realistic gain from combination therapy comes from the
extremely low probability of simultaneous emergence of both
resistances in the same cell (joint mutation rate $10^{-18}$
versus $10^{-9}$ for single mutation), which suppresses the
\emph{occurrence} of resistance more than it stretches the
\emph{time} to fixation once a doubly resistant cell exists.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a vaccination programme for measles in a population of
$10^7$ that starts at vaccination coverage 60\%. State the
target coverage, the campaign duration, the prioritisation by age
class, and the expected case load during the campaign. Use $R_0 =
15$ and waning immunity time scale 50 years.

\paragraph{Discussion.} Herd-immunity threshold $p_c = 14/15
\approx 0.933$; the population is 33 percentage points short. A
campaign delivering one million additional vaccinations per year
takes 3.3 years to close the gap, during which $R_t > 1$ and
outbreaks remain possible. Prioritisation: children 6--18 months
(the post-maternal-antibody window during which routine
vaccination has the highest per-dose effect) and rural
under-immunised cohorts (the standard equity-and-coverage joint
target). Expected case load over the campaign: with $R_t \approx
6$ at the start (effective reproduction number with 60\%
immunity), small outbreaks of 10--500 cases are likely in
under-vaccinated pockets; an unmitigated outbreak in a fully
susceptible sub-population would reach the herd-immunity fraction
within 4--6 weeks (Exercise 14 above). Waning immunity at the
50-year timescale implies a per-year loss of $\sim 2\%$ coverage,
so maintenance vaccination at $\sim 200{,}000$ per year is
required to sustain the threshold once reached.
\end{exercise}

\begin{exercise}
Design a captive-breeding programme for a species with current
population $N = 30$ in the wild and estimated Allee threshold
$A = 50$. State the programme size, the release schedule, the
genetic-management plan, and the criterion for declaring the wild
population self-sustaining.

\paragraph{Discussion.} Target wild population of $\sim 100$
($2 \times A$) to provide buffer against demographic stochasticity.
Captive programme of $\sim 50$ founders drawn to represent the
wild gene pool's allele frequencies; effective population size
$N_e \geq 50$ to prevent short-term inbreeding depression.
Release schedule: $20$--$30$ animals per year over 5--7 years to
lift the wild population through the Allee threshold while
avoiding shock-release mortality. Genetic management: mean
kinship minimisation in the captive programme; periodic
infusion of wild-caught individuals back into the captive stock
to maintain heterozygosity. Self-sustaining criterion: $\geq 3$
consecutive years of wild population $> A$ with positive observed
growth rate and reproductive contribution from $> 50\%$ of adults
(not just a few dominant breeders). The black-footed ferret
programme (1985--present) and the California condor programme
(1987--present) followed this template; the kakapo programme
applied the same logic with more intensive per-individual
management.
\end{exercise}

\begin{exercise}
Design a marine reserve network for a coastal fishery. State the
number, size, and spacing of reserves, the larval-dispersal
parameters that constrain the design, and the metric you will use
to evaluate the network's effectiveness over a 20-year horizon.

\paragraph{Discussion.} Larval-dispersal distance is the load-bearing
parameter; for many coastal teleosts the mean larval-dispersal
distance is 10--50\,km. Network design: 8--12 reserves, each
$\sim 20$--$50\,\text{km}^2$ ($\sim 5\%$--$10\%$ of total fishery
area), spaced 30--100\,km apart so larvae from one reserve seed
adjacent fished areas. The metric is twofold: (1) spillover index
(observed catch-per-unit-effort in fished areas adjacent to
reserves versus control areas distant from reserves); (2)
reserve-internal biomass density relative to baseline. Twenty-year
horizon: expect 2--5x biomass increase inside reserves within
5--10 years, and a measurable spillover signal (10\%--30\% CPUE
increase in adjacent fished areas) within 10--15 years; over the
20-year horizon, the cumulative catch outside reserves should
match or exceed pre-reserve catch from the same area. The
Mediterranean Posidonia-seagrass network and the Great Barrier
Reef rezoning of 2004 provide the empirical record for design
parameter calibration.
\end{exercise}

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
A 2020 COVID-19 model report predicts a peak of 100{,}000 daily
cases in a country, with a 95\% confidence interval of (30{,}000,
350{,}000). The observed peak was 250{,}000 daily cases. Was the
model wrong? State criteria for what ``wrong'' means in this
context and apply them to the prediction.

\paragraph{Discussion.} ``Wrong'' has at least three meanings here.
(1) Calibration: the 95\% CI contains the observation (250k is
within [30k, 350k]), so by calibration the model was \emph{not}
wrong. The model accurately quantified its uncertainty. (2)
Point-prediction accuracy: the point estimate (100k) is off by a
factor of 2.5; if the report's headline was the point estimate
without the interval, the report misled. (3) Structural validity:
asking whether the assumptions held independently of whether the
prediction came in. A model that calibrated correctly by accident
(point estimate wrong, CI right by virtue of structural error) is
weaker than a model that was correctly calibrated for the right
reasons. The engineering answer is to read all three: (1) and (3)
pass; (2) reflects an over-confident headline that the modelling
team should have anticipated.
\end{exercise}

\begin{exercise}
A regulator proposes that all monotherapy antibiotic prescribing be
replaced by combination therapy for one named pathogen. State three
classes of evidence that would support or undermine the proposal and
the timescale on which each would become available.

\paragraph{Discussion.} Three evidence classes. (1) Clinical
outcomes: comparative trials of combination versus monotherapy on
cure rates, time-to-defervescence, and treatment-failure incidence.
Timescale: 2--5 years from trial design to reporting; meta-analytic
synthesis later. (2) Surveillance data on resistance prevalence:
year-on-year tracking of resistance frequencies in clinical
isolates under the new regimen versus historical baseline.
Timescale: 1--2 years for first signal, 5--10 years for confident
trend assessment given the typical $\sim 1\%$/year resistance
drift in single-mutation systems. (3) Side-effect and adherence
data: combination therapy doubles drug exposure and may cut
adherence, both of which reduce real-world effectiveness; pharmacy
records and clinical-trial side-effect reports inform this.
Timescale: 6 months to 2 years. The regulator should require all
three before scaling the proposal.
\end{exercise}

\begin{exercise}
A conservation organisation proposes that a population of
$N = 60$ individuals in the wild can be allowed to recover without
captive breeding intervention, on the basis that the per-capita
growth rate is positive. State the additional information you
would require to assess the proposal and the threshold at which
your judgement would shift toward intervention.

\paragraph{Discussion.} Required information: (1) effective
population size $N_e$ (typical vertebrate ratio $N_e/N \approx
0.1$--$0.3$ implies $N_e = 6$--$18$, well below the
$N_e \geq 50$ short-term inbreeding-prevention threshold); (2)
estimated Allee threshold $A$ from breeding biology; (3)
year-to-year variance $\sigma_r^2$ in the per-capita growth rate
(positive mean is not enough; the geometric long-term rate is
$\bar r - \sigma_r^2/2$, which can be negative for variable
populations); (4) demographic-stochasticity-induced extinction
probability over a 100-year horizon (population viability
analysis). Threshold for intervention: any of (a) $N_e < 50$, (b)
$N < 2A$, (c) projected extinction probability $> 5\%$ over 100
years, (d) observed per-capita growth rate confidence interval
includes zero. Under any single trigger, captive-breeding
intervention is recommended; the IUCN Red List criteria operationalise
roughly this framework.
\end{exercise}

\begin{exercise}
A fisheries manager proposes that the harvest rate of a stock be
set to the maximum sustainable yield, computed from the logistic
model with $K$ estimated from current biomass. State the principal
risks of this proposal and the additional precautionary buffer you
would recommend.

\paragraph{Discussion.} Principal risks. (1) $K$ estimated from
current biomass is path-dependent: if current biomass is below
the historical long-run $K$ (the baseline-shift problem), the
calculated MSY is below the true MSY but the stock dynamics see
a higher harvest pressure than the model predicts. (2) The
logistic model assumes a fixed $K$; ecosystem regime shifts
(temperature, prey availability, predation) move $K$ on
multi-decadal timescales. (3) MSY-target harvest sits at the
unstable equilibrium $N = K/2$: any negative recruitment shock or
$r$ over-estimate drops the population into the unstable basin
and collapse becomes deterministic. The cod precedent (§8.7) is
direct evidence of this failure mode. Recommended buffer: harvest
at $0.6$--$0.75$ of MSY (not $1.0$); recompute $K$ annually from
direct biomass survey rather than from catch records; trigger a
moratorium if stock biomass falls below $0.4 K$. The precautionary
buffer is the difference between a stock that recovers from a
mistake and one that collapses.
\end{exercise}

\begin{exercise}
A modelling group reports that the herd-immunity threshold for a new
respiratory virus is 70\%, based on an SIR model with homogeneous
mixing and an $R_0$ of 3. State three structural assumptions the
model makes that may shift the true threshold by tens of
percentage points, and state the direction of each shift.

\paragraph{Discussion.} (1) Homogeneous mixing: the assumption that
every susceptible is equally likely to contact every infectious.
Heterogeneity lowers the threshold because high-contact individuals
are infected first and removed from the susceptible pool, leaving a
disproportionately low-contact remainder. Direction: lower (often
from 70\% to 40\%--50\% under realistic contact-rate variance).
(2) Permanent immunity on recovery: the assumption that recovered
individuals stay immune. Waning immunity raises the effective
threshold because the immune pool shrinks over time and the threshold
must be maintained dynamically. Direction: higher, by 5--15 percentage
points depending on the waning rate. (3) Constant $R_0$: the
assumption that the basic reproduction number does not change.
Variant emergence (Alpha, Delta, Omicron during SARS-CoV-2)
raises $R_0$ and the threshold along with it. Direction: higher,
sometimes by 20--30 percentage points (the gap between
ancestral-SARS-CoV-2 threshold $\approx 0.65$ and Omicron-period
threshold $\approx 0.90$).
\end{exercise}

\begin{exercise}
A public-health authority is choosing between two interventions for
an outbreak with $R_0 = 2$: vaccinate 50\% of the population over six
months, or impose contact restrictions that reduce $\beta$ by 60\%
for two months. Estimate the final attack rate under each scenario
and identify two factors not in the SIR model that would change
your choice.

\paragraph{Discussion.} Vaccination scenario: 50\% coverage drives
$R$ from 2.0 to $R_0 (1 - 0.5) = 1.0$. The post-vaccination wave
just-barely passes herd immunity threshold $p_c = 1 - 1/R_0 =
0.5$, so the residual epidemic is at the edge of self-extinction.
Final attack rate among unvaccinated: depends on speed of rollout
versus growth; if vaccination is slow (uniform over 6 months) the
unvaccinated cohort experiences a sub-critical wave with attack
rate $\sim 20$\% (from the final-size relation at $R_0 = 1$). With
vaccinated population, total attack rate $\sim 10$\% of population.
Contact-restriction scenario: $\beta$ falls by 60\% to give
effective $R = 0.4 \cdot R_0 = 0.8$, sub-critical, so the outbreak
contracts during the 2-month period. Restoring contact rates after
2 months returns $R$ to 2.0; final attack rate depends on the
remaining susceptible pool. If most of the pre-restriction
infections were avoided, the post-restriction wave runs the
final-size relation at $R_0 = 2$ over a still-susceptible
population, giving $\sim 80$\% final attack rate of remainder.

Factors outside the SIR model that change the choice: (1)
vaccination side effects, infrastructure, equity of distribution
(SIR is structureless on who gets vaccinated first); (2)
contact-restriction economic and mental-health costs; (3) the
emergence of a new variant during the 8-month total window
(changes $R_0$ and may render either intervention obsolete).
A working public-health authority combines the two: vaccinate
priority groups first while contact restrictions buy time, then
release restrictions as coverage rises. Neither isolated
intervention is a good plan.
\end{exercise}
