\chapter{CAD and engineering drawing}
\label{vol08:ch05}

\epigraph{The drawing is the contract.}{anonymous mechanical engineer}

\chapmeta{Half-life: medium-life. Archetypes: interface, failure. Exercise target: 30. Project track: physical. Named cases: Mars Climate Orbiter.}

\noindent
An engineering drawing is a contract between the designer and the manufacturer. It states what shape, what dimension, what tolerance, what surface finish, what material, what processing, what marking. A drawing's purpose is to be unambiguous; a manufacturer reading the drawing and a different manufacturer reading the same drawing should produce parts that interchange with each other. The discipline that makes this possible is engineering drawing, modernised as 3D parametric CAD with GD\&T (Geometric Dimensioning and Tolerancing) annotations. The standard references are ASME Y14.5 (the GD\&T standard) and Krulikowski \cite{text:krulikowski-gdt}.

\section{Engineering drawing as communication}\pathtag{core}

A working engineering drawing has several components:

\textbf{Views.} Orthographic projections (front, top, side) showing the part's geometry. Modern practice favours first-angle projection (Europe, Asia) or third-angle projection (US). Auxiliary views, section views, and detail views as needed.

\textbf{Dimensions.} Numerical values for sizes and locations.

\textbf{Tolerances.} The allowable variation on each dimension.

\textbf{Surface-finish specifications.} The roughness allowed for each surface.

\textbf{Material specification.} The material grade, condition, treatment.

\textbf{Notes.} Special manufacturing or inspection requirements.

\textbf{Title block.} Drawing number, revision, date, drafter, approver, scale, units.

\textbf{Revision history.} Each modification with date and reason.

The drawing's purpose is to be readable by someone who has never met the designer and will not have the opportunity to ask questions. Modern engineering organisations enforce drawing standards (company-specific or industry-standard) to ensure consistency.

\section{Tolerance: dimensional, geometric (GD\&T)}\pathtag{core}

A part cannot be manufactured to the nominal dimension exactly. The \engterm{tolerance} specifies the allowable variation. The engineer's discipline is to set tolerances tight enough that parts function and interchange, loose enough that manufacturing is economical.

\subsection{Dimensional tolerance}

A dimension of $25\,\si{\milli\meter}$ might be specified as $25 \pm 0.1\,\si{\milli\meter}$ (bilateral), $25\,^{+0.0}_{-0.2}\,\si{\milli\meter}$ (unilateral), or as a fit class ($25H7$ for a hole at the standard tolerance grade 7).

The relationship between mating parts is the \engterm{fit}: clearance fit (parts move freely), transition fit (uncertain whether they have clearance or interference), interference fit (require force or thermal expansion to assemble). The ISO fits system tabulates standard combinations: H7/g6 for a sliding fit, H7/n6 for a tight transition, H7/p6 for a light press fit.

\subsection{Geometric tolerances (GD\&T)}

Dimensional tolerances alone cannot specify all the requirements. A shaft might be cylindrical to size but not straight; two faces might be flat but not parallel; a hole might be the right size but not centred. GD\&T fills these gaps with explicit specifications of form, orientation, location, runout.

The 14 GD\&T characteristics (ASME Y14.5):

\textbf{Form (4):} straightness, flatness, circularity, cylindricity.

\textbf{Orientation (3):} perpendicularity, parallelism, angularity.

\textbf{Location (3):} position, concentricity, symmetry.

\textbf{Runout (2):} circular runout, total runout.

\textbf{Profile (2):} profile of a line, profile of a surface.

Each is specified by a control frame on the drawing: the characteristic symbol, the tolerance value, the datum reference. A datum is a feature the part is referenced against (a flat face, an axis, a centre plane).

\subsection{Tolerance stack-up}

The \engterm{tolerance stack-up} is the accumulated tolerance across multiple dimensions or parts. The classical worst-case method sums the individual tolerances; the statistical (RSS) method takes the root-sum-square. A shaft passing through a series of bearings has positional tolerance from each bearing's position and from the shaft's own runout; the cumulative tolerance is what the assembly sees.

The engineering discipline is to identify the critical-functional dimensions and compute the stack-up explicitly. A six-tolerance chain at $\pm 0.1\,\si{\milli\meter}$ each has a worst-case stack of $\pm 0.6\,\si{\milli\meter}$, an RSS stack of $\pm 0.24\,\si{\milli\meter}$. The choice between worst-case and RSS depends on the assembly criticality.

\section{Surface finish}\pathtag{standard}

A \engterm{surface finish} (also called surface roughness, surface texture) specifies how rough the manufactured surface may be. The classical metric is $R_a$ (arithmetic average roughness), with $R_z$ (peak-to-valley) and others as supplementary.

Typical $R_a$ values:
\begin{tabular}{lll}
\toprule
$R_a$ range ($\si{\micro\meter}$) & Typical achievement \\
\midrule
$> 6.3$ & rough machining; sand casting \\
$1.6$-$6.3$ & general machining; investment casting \\
$0.4$-$1.6$ & finish machining; grinding \\
$0.1$-$0.4$ & precision grinding; honing \\
$< 0.1$ & lapping; super-finishing \\
\bottomrule
\end{tabular}

The engineer specifies surface finish on the drawing where it matters: bearing seats, sealing surfaces, sliding contacts. Specifying tighter than necessary adds manufacturing cost without function; specifying looser produces functional failure.

\section{Parametric CAD: features, constraints}\pathtag{standard}

Modern engineering design uses \engterm{parametric CAD}: the part is built from features (extrudes, revolves, holes, fillets) that are parameterised. Changing a parameter (a length, an angle, a radius) updates the model and all derived drawings.

\subsection{Features and constraints}

A typical CAD model is a sequence of features:
\begin{enumerate}[leftmargin=*]
  \item A sketch on a plane (2D geometry with dimensional and geometric constraints).
  \item An operation (extrude the sketch into 3D, revolve it, sweep along a curve).
  \item Modifying operations (fillet edges, chamfer corners, shell out walls, draft for moulding).
  \item Holes, threads, patterns.
\end{enumerate}

The constraints are critical. A well-constrained sketch has zero degrees of freedom: every dimension and relationship is specified. An under-constrained sketch deforms under parameter changes; an over-constrained sketch produces conflicts.

\subsection{Modelling discipline}

\textbf{Design intent.} Build the model so that the parameters reflect the design intent. A hole's position should be defined from the feature that drives it, not from an arbitrary edge.

\textbf{Feature ordering.} Place coarse features first, fine details last. Place stable features (those unlikely to change) before features that depend on them.

\textbf{Reusable references.} Use construction geometry (axes, planes, points) as stable references.

\textbf{Naming.} Name features and parameters meaningfully. A model with $50$ features and default names (``Boss-Extrude1'' through ``Boss-Extrude50'') is unmaintainable.

\subsection{The major CAD families}

\textbf{Mainstream MCAD.} SolidWorks, Inventor, Solid Edge, Creo. Mid-range; widely used in industry.

\textbf{High-end MCAD.} NX, CATIA. Used in aerospace, automotive OEMs. Capable of handling very large assemblies and complex surfaces.

\textbf{Direct modelling.} Onshape, Fusion 360. Mix parametric and direct; cloud-collaborative.

\textbf{Programmatic CAD.} OpenSCAD, CadQuery, FreeCAD Python scripting. Used by hobbyists and for parametric-by-code workflows.

\section{Assembly and motion}\pathtag{standard}

A CAD \engterm{assembly} combines individual parts into a higher-level model. The assembly's constraints (mates) specify how the parts fit: coincident faces, concentric axes, parallel planes, distance offsets.

\subsection{Assembly mate strategy}

Modern CAD systems offer dozens of mate types. The engineering discipline is to use the smallest number of mates that fully constrain the assembly. Over-constrained assemblies are fragile to parameter changes; under-constrained assemblies have unintended degrees of freedom.

\subsection{Motion studies}

Modern CAD systems include motion-study capabilities: drive a parameter (an angle, a position) and observe how the assembly moves. Used for verifying clearances, animating mechanisms, generating cam profiles, exporting kinematic data for downstream analysis.

\section{PLM and BOM}\pathtag{standard}

\subsection{Product lifecycle management}

\engterm{Product lifecycle management} (PLM) is the software discipline of managing engineering data across a product's life: CAD files, drawings, specifications, BOMs, change orders, supplier data, configuration, deployment. PLM systems (Windchill, Teamcenter, Enovia, Aras) provide version control, workflow management, access control, and integration with ERP and MES systems.

The engineering value: PLM ensures that everyone working on a product is referring to the same version of every component. PLM's failures are routine: configuration drift, version-history loss, broken cross-references between systems.

\subsection{Bill of materials}

A \engterm{bill of materials} (BOM) lists every part in an assembly: part number, description, quantity, source, material, drawing reference. The BOM is the working document for procurement, manufacturing, and assembly.

A modern BOM is hierarchical: a top-level assembly contains sub-assemblies, each with its own BOM. Modern engineering practice generates the BOM directly from the CAD assembly to ensure consistency.

\section{Drawing for manufacture}\pathtag{standard}

\subsection{Manufacturing-driven drawing}

The drawing serves the manufacturer. Specifications that make manufacturing easier (loose tolerances where function permits, standard hole sizes, common surface finishes) reduce cost; specifications that make manufacturing harder (tight tolerances, exotic finishes, non-standard hole sizes) increase cost.

The engineering discipline is to draw for the intended manufacturing route. A part to be CNC-milled has different drawing requirements than one to be cast or 3D-printed.

\subsection{Drawing conventions}

\textbf{Title block.} Identifies the drawing.

\textbf{Default tolerances.} A note in the title block specifies tolerances for unspecified dimensions (e.g., ``$\pm 0.5\,\si{\milli\meter}$ unless otherwise specified'').

\textbf{Material and treatment.} Specified in the title block or in a note.

\textbf{Surface finish.} Default in the title block, with specific values called out where needed.

\textbf{Notes.} Special manufacturing requirements (e.g., ``Polish weld smooth before inspection,'' ``Apply X coating after machining'').

\textbf{Inspection requirements.} What measurements the manufacturer must make on each part.

\section{Failure: tolerance-stack errors, drawing ambiguity, unit-system clash}\pathtag{core}

\subsection{Tolerance stack-up errors}

A part designed without explicit tolerance analysis can fail at assembly even though every individual part is within tolerance. The classical case: a chain of features whose individual tolerances of $\pm 0.1\,\si{\milli\meter}$ accumulate to $\pm 0.6\,\si{\milli\meter}$ at the critical interface, exceeding the function's tolerance budget. The engineering response: compute the stack-up at design time, specify tighter tolerances on critical-path features.

\subsection{Drawing ambiguity}

A drawing that admits two interpretations produces parts that vary depending on which interpretation the manufacturer chose. Common patterns: missing datum reference (perpendicularity to what?); ambiguous fillet (radius or chord-length?); undimensioned feature (assumed to match a CAD model that the manufacturer does not have); contradiction between two views.

The engineering response is rigorous drawing review by a second engineer. The principle: a drawing should be readable by a competent manufacturer in another country, in another language community, working from the drawing alone.

\subsection{Unit-system clash: Mars Climate Orbiter}

The Mars Climate Orbiter case \cite{acc:nasa-mco-mib}, named in Chapter 1 and Volume I, is the canonical drawing-ambiguity failure at industrial scale. The interface specification was written in metric units; a sub-contractor's software produced data in imperial units. The interface was unambiguous as a written document; the unit-system handoff was not enforced at the data-handoff point. The cumulative error destroyed the spacecraft.

The engineering response: every drawing and every interface specification carries an explicit unit notation (``All dimensions in millimetres unless otherwise specified''); every data interface enforces unit compatibility at the boundary (a parameter passed between two systems carries its units, and the receiving system rejects mismatches).

\begin{principle}[The drawing is the source of truth]
A part is what the drawing specifies, not what the designer intended. A drawing that does not specify what the designer intended produces parts that diverge from the intention. The engineering discipline is to make the drawing the unambiguous, complete, audited statement of the part: every dimension, every tolerance, every surface finish, every material, every treatment, every inspection. The drawing's quality is the manufacturer's only window into the designer's mind.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[Produce a fully-detailed drawing pack for a small assembly]
Pick a small mechanical assembly (a hand tool, a bracket assembly, a small mechanism). Produce a complete drawing pack: 3D CAD models of all parts, 2D drawings with GD\&T, assembly drawing, BOM, notes for manufacturing.

\begin{enumerate}[leftmargin=*]
  \item Identify the parts (3 to 7 individual parts plus standard fasteners).
  \item Model each part in your CAD system. Apply design intent: name features meaningfully, dimension to drive the design.
  \item Create 2D drawings for each part: views, dimensions, GD\&T tolerances, surface-finish callouts, material spec, notes.
  \item Create the assembly drawing with section views as needed.
  \item Generate the BOM from the assembly.
  \item Have a second engineer review the drawings for ambiguity, completeness, and manufacturability.
\end{enumerate}

The deliverable is the CAD files, the drawing pack (PDF), the BOM, and a short report (no more than four pages) on the engineering decisions and the issues the review uncovered.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
For a $30H7/g6$ fit, look up the actual hole and shaft tolerances. Compute the clearance range.
\end{exercise}

\begin{exercise}
A chain of 5 dimensions, each $\pm 0.1\,\si{\milli\meter}$, has worst-case stack of $\pm 0.5\,\si{\milli\meter}$. Compute the RSS stack.
\end{exercise}

\begin{exercise}
A surface-finish specification calls for $R_a \leq 0.8\,\si{\micro\meter}$. State the typical manufacturing process required.
\end{exercise}

\begin{exercise}
For a $20\,\si{\milli\meter}$ diameter shaft passing through two bearings $200\,\si{\milli\meter}$ apart, with bearing concentricity tolerance $\pm 0.05\,\si{\milli\meter}$ each, compute the resulting axial misalignment angle.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the relationship between worst-case and RSS tolerance stacks: when does the RSS method underestimate the actual stack?
\end{exercise}

\begin{exercise}
Show that an under-constrained CAD sketch has degrees of freedom that lead to ambiguous geometry when parameters change.
\end{exercise}

\begin{exercise}
Derive the relationship between fit class (H7/g6, H7/p6) and the assembled function (sliding, interference).
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the cost ratio of $\pm 0.01\,\si{\milli\meter}$ vs.\ $\pm 0.1\,\si{\milli\meter}$ tolerance on a CNC-milled part.
\end{exercise}

\begin{exercise}
Estimate the surface-finish-class cost difference between $R_a = 6.3\,\si{\micro\meter}$ and $R_a = 0.4\,\si{\micro\meter}$.
\end{exercise}

\begin{exercise}
Estimate the number of dimensions and tolerances on a moderate-complexity machined part.
\end{exercise}

\begin{exercise}
Estimate the design time vs.\ tolerance-analysis time ratio for a precision assembly.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Build a parametric CAD model of a part with a deliberate under-constrained sketch. Change parameters; observe the model's failure modes.
\end{exercise}

\begin{exercise}
Build an assembly with five parts; apply mates to fully constrain. Verify by attempting to move each part.
\end{exercise}

\begin{exercise}
Use a tolerance-stack analysis tool (or build one) to compute the cumulative tolerance through a five-part assembly. Compare worst-case to RSS.
\end{exercise}

\begin{exercise}
Use a motion study to verify that a four-bar linkage's coupler curve matches the kinematic prediction of Chapter 1.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design the tolerance budget for an assembly that must achieve $\pm 0.2\,\si{\milli\meter}$ at a critical mating interface across five contributing dimensions.
\end{exercise}

\begin{exercise}
Design the drawing for a machined part requiring $\pm 0.05\,\si{\milli\meter}$ on a critical bore and $\pm 0.5\,\si{\milli\meter}$ on cosmetic dimensions.
\end{exercise}

\begin{exercise}
Design the GD\&T scheme for a part with three datums and four geometric tolerance specifications.
\end{exercise}

\begin{exercise}
Design the inspection plan for a precision part: which measurements at what frequency.
\end{exercise}

\subsection*{Judgement}\pathtag{core}

\begin{exercise}
The Mars Climate Orbiter case \cite{acc:nasa-mco-mib} hinged on a unit-system handoff. State three engineering practices that would have caught the error at the drawing or interface-specification level.
\end{exercise}

\begin{exercise}
A team's CAD model has 200 features and default names. State the engineering risks and the remediation.
\end{exercise}

\begin{exercise}
A team uses GD\&T inconsistently across drawings; some have explicit datum references, others rely on default. State the manufacturing risks.
\end{exercise}

\begin{exercise}
A team's tolerance analysis uses RSS exclusively. State the assumptions and the failure mode if they are violated.
\end{exercise}

\begin{exercise}
A team uses ISO standard fits (H7/g6, etc.) throughout. State the engineering benefits and the limits of the practice.
\end{exercise}

\begin{exercise}
A team's drawing review is done by the original designer only. State the engineering risk.
\end{exercise}

\begin{exercise}
A team's PLM system has multiple ``approved'' versions of the same drawing in different projects. State the engineering risk.
\end{exercise}

\begin{exercise}
A senior engineer argues that ``modern CAD makes 2D drawings obsolete; the 3D model is the source of truth.'' Critique. State the workflows where 2D drawings remain essential.
\end{exercise}

\begin{exercise}
A team specifies $R_a \leq 0.2\,\si{\micro\meter}$ for every surface of every part. State the manufacturing-cost implication.
\end{exercise}

\begin{exercise}
A team's drawing is ambiguous on the location of a hole. The manufacturer chose one interpretation; the design intended the other. State the engineering remediation.
\end{exercise}

\begin{exercise}
A team uses a free-form CAD direct-modelling tool for engineering drawings. State the engineering risks compared to parametric CAD.
\end{exercise}
