% Volume 3: Force
% Opener: tagline + scope. Sits immediately after the \part{} declaration.

\vspace*{1.5cm}
\begin{center}
\itshape\large
The grammar of why things move and don't
\end{center}
\vspace{0.5cm}

\noindent
We point Volume II's mathematics at the mechanical world: statics, dynamics, mechanics of deformable solids, fluid mechanics through compressible flow.

Force is how the mechanical world settles its accounts. A structure
stands because the forces on it sum to zero; a machine moves because
they do not; a part fails because a force found a path its designer
never priced. Volume I taught the reader to measure, and Volume II gave
measured quantities mathematical form as functions, derivatives,
integrals, differential equations, and linear systems. Volume III spends
both. Every chapter here opens with a free-body diagram, a rate of
change, a conservation statement, or a matrix of coefficients, and the
reader who did the work of the first two volumes will recognise the
instrument before the mechanics renames it.

The volume moves from the rigid to the deformable to the flowing. It
begins with forces and the free-body diagram (Chapter 1) and the
equilibrium of rigid bodies (Chapter 2), the discipline of accounting
for every force before writing a single equation. It then lets bodies
deform: stress and strain (Chapter 3), the internal bookkeeping that
decides whether a member yields, and the energy methods (Chapter 5) that
turn deformation into a scalar often easier to work than the forces
themselves. Dynamics (Chapter 4) sets the rigid body in motion;
vibrations (Chapter 6) and acoustics (Chapter 7) follow motion into
oscillation and oscillation into wave. Chapters 8 and 9 return to
structures that carry load by bending and by membrane action: beams,
columns, frames, plates, and shells. The last four chapters open the
fluid world: statics (Chapter 10), inviscid flow (Chapter 11), the
viscous boundary layer (Chapter 12), and compressible flow to the shock
(Chapter 13).

Four archetypes recur. \emph{Equilibrium}: the sum of forces and moments
on a body at rest is zero, and the reader who can draw the free-body
diagram owns half the problem before any equation. \emph{Conservation}:
mass, momentum, and energy cross the boundary of a control volume in
amounts that must balance, and the same balance equation returns in
Volume IV for energy and Volume V for reacting species.
\emph{Failure}: every chapter names the mechanism by which its
structures give way, from yield and buckling to fatigue, resonance,
cavitation, and shock. \emph{Scaling}: a beam's stiffness grows as the
cube of its depth, a column's critical load falls as the square of its
length, and drag climbs as the square of speed, so the engineer who
carries these exponents reads a design before computing it.

By the end of the volume the reader can take a mechanical system, draw
its free-body diagram, write its equilibrium or its equation of motion,
size its members against yield and buckling, estimate its natural
frequencies, and trace a fluid across it from the static head to the
compressible limit. The reader can also name how each of those will
fail, and price the failure.

The mathematics of Volume II is the working language throughout. Statics
is a linear system; dynamics is an ordinary differential equation; a
vibrating structure is an eigenvalue problem; a stress state is a
symmetric matrix whose eigenvalues are its principal stresses; a flow
field is a vector calculus of divergence and circulation. Where a
chapter needs a tool the reader has not met, it says so and points back.
The forces are new; the grammar is the one the reader already owns.

\vspace{1cm}
