\chapter{Metals and alloys}
\label{vol05:ch05}

\epigraph{The plastic flow of crystalline solids is the motion of
dislocations through the lattice; their density, their interaction with
each other and with obstacles, sets the macroscopic strength.}{The
modern statement of the dislocation theory of plasticity, articulated
in \textcite[ch.~5]{text:dieter-mechanical-metallurgy}.}

\chapmeta{Half-life: foundational for the metallurgy of crystal structure and dislocations; medium-life for the catalog of commercial alloys (compositions and properties update with each generation). Archetypes: scaling, failure. Exercise target: 30. Project track: identification of an unknown metal sample by household testing. Hazard class: low (density measurement, hand-held hardness comparison, conductivity probe). Reader path default: standard, with sections explicitly tagged. Named cases: none required; weld, hydrogen-embrittlement, and stress-corrosion-cracking failures are described by mechanism.}

\noindent
Metals are the structural backbone of the engineered world. The
chapter develops the crystal structures of the metallic elements, the
dislocation theory of plastic deformation that bridges the atomic
bonding of Chapter~1 to the macroscopic strength of structural
components, the four mechanisms (solid-solution, work-hardening,
precipitation, grain-size) by which alloying and processing strengthen
metals, the families of engineering metals (steels, cast irons,
aluminium, copper, titanium, magnesium, nickel-based superalloys),
the joining of metals by welding, and the failure modes specific to
metallic structural components: weld defects, hydrogen embrittlement,
and stress-corrosion cracking. The chapter is the working-engineer
companion to the phase diagrams of Chapter~4 and the corrosion
mechanisms of Chapter~9.

\section{Crystal structures of metals}\pathtag{core}

The metallic elements crystallise predominantly in three lattice
types: face-centred cubic (FCC), body-centred cubic (BCC), and
hexagonal close-packed (HCP). The three lattices were introduced in
Chapter~1 of this volume; the present section names which metals
occupy which lattice and develops the slip systems that each lattice
makes available for plastic deformation.

\engterm{FCC metals}. Aluminium, copper, gold, silver, nickel,
platinum, lead, and $\gamma$-iron (austenite, above $912\,\degC$).
The lattice has packing fraction $0.74$, the maximum for equal
spheres, and twelve nearest neighbours. The close-packed planes are
the $\{111\}$ planes (four equivalent planes); the close-packed
directions are the $\langle 110 \rangle$ directions (three within
each $\{111\}$ plane). The product gives twelve $\{111\}\langle 110
\rangle$ slip systems, the largest number of any common metal lattice.
The abundance of slip systems is why FCC metals are typically ductile:
under any applied stress state, a slip system that is favourably
oriented for dislocation motion is available.

\engterm{BCC metals}. $\alpha$-iron (ferrite, below $912\,\degC$),
chromium, molybdenum, tungsten, vanadium, sodium, potassium, niobium,
tantalum, and the $\delta$-iron above $1394\,\degC$. The lattice has
packing fraction $0.68$ (not close-packed) and eight nearest
neighbours plus six second-nearest at $15\,\%$ greater distance. The
close-packed directions are the $\langle 111 \rangle$ directions. The
slip planes are less clear-cut than in FCC; the primary slip systems
are $\{110\}\langle 111 \rangle$ (twelve systems), with additional
slip on $\{112\}\langle 111 \rangle$ (twelve more) and on
$\{123\}\langle 111 \rangle$ (twenty-four more) at higher temperatures
and stresses. BCC metals are less ductile than FCC at low temperatures
and exhibit a ductile-to-brittle transition temperature (DBTT) above
which plastic deformation is possible and below which the material
fractures brittly. The DBTT depends strongly on alloy composition and
microstructure; for plain-carbon steels it is in the $-50$ to $+50\,
\degC$ range, with severe consequences for ships and pipelines
operating in cold climates (the brittle fracture of Liberty ships in
the North Atlantic during World War II is the canonical historical
example).

\engterm{HCP metals}. Zinc, magnesium, titanium ($\alpha$-Ti below
$882\,\degC$), zirconium, beryllium, cadmium, cobalt ($\alpha$-Co).
The lattice has the same packing fraction as FCC ($0.74$) but stacking
sequence ABABAB rather than ABCABC. The close-packed plane is the
basal $(0001)$ plane; the close-packed directions are the
$\langle 11\bar{2}0 \rangle$ directions. The basal slip system
provides only three slip systems, and at room temperature it is the
only easily activated system; HCP metals are consequently anisotropic
and less ductile than FCC. The $c/a$ ratio (the ratio of the
hexagonal unit cell's $c$-axis to $a$-axis) is $1.633$ for ideal HCP
(spheres in contact); zinc has $c/a = 1.856$ and magnesium $c/a =
1.624$, the deviation from the ideal corresponding to non-spherical
electronic shapes of the atoms.

\engterm{Other structures}. Manganese has a complex cubic structure;
mercury at room temperature is liquid; uranium has a base-centred
orthorhombic $\alpha$ phase below $660\,\degC$. The remaining metals'
structures are catalogued in standard handbooks
\cite[ch.~1]{text:smallman-physical-metallurgy}.

\engterm{The slip-system count and ductility}. The von Mises criterion
for a polycrystalline metal to deform plastically by slip is that
five independent slip systems must be available. FCC metals have
twelve and easily satisfy the criterion. BCC metals nominally have
twelve $\{110\}\langle 111 \rangle$ systems but only five are
independent; the criterion is just satisfied. HCP metals have three
basal slip systems, plus prismatic and pyramidal systems at higher
stresses; the von Mises criterion is not satisfied by basal slip
alone, and HCP polycrystals therefore exhibit limited room-temperature
ductility unless twinning or non-basal slip is activated. The
engineering consequence is that wrought HCP alloys are typically
processed at elevated temperatures where non-basal slip is easier and
that HCP castings are coarser-grained and less ductile than
comparable FCC or BCC castings.

\section{Plastic deformation and dislocations}\pathtag{core}

A perfect crystal subjected to a shear stress would deform by the
collective sliding of one atomic plane over another. The theoretical
shear strength of such a perfect crystal is $\tau_{\text{theor}} \sim
G/(2\pi)$, where $G$ is the shear modulus. For aluminium ($G \approx
26\,\text{GPa}$), $\tau_{\text{theor}} \sim 4\,\text{GPa}$. The
measured shear yield strength of pure aluminium is $\sim 7\,\text{MPa}$,
a factor of $\sim 600$ lower. The gap is the central puzzle of
twentieth-century metallurgy, resolved in 1934 with the dislocation
hypothesis advanced independently by Taylor, Orowan, and Polanyi: real
crystals contain line defects (dislocations) that move under shear
stress at a fraction of the theoretical shear strength
\cite[ch.~6]{text:smallman-physical-metallurgy}.

\engterm{The Burgers vector and dislocation types}. A
\engterm{dislocation} is characterised by the closure failure of a
Burgers circuit drawn around the dislocation core, the
\engterm{Burgers vector} $\vec{b}$. The vector's direction relative
to the dislocation line distinguishes the two basic types: an
\engterm{edge dislocation} has $\vec{b}$ perpendicular to the line
(geometrically, an extra half-plane of atoms is inserted into the
crystal, ending at the dislocation line); a \engterm{screw
dislocation} has $\vec{b}$ parallel to the line (geometrically, the
lattice planes wind helically around the dislocation). Mixed
dislocations have $\vec{b}$ at an intermediate angle. The Burgers
vector is invariant along the dislocation line (it is a topological
invariant of the closure failure); the line itself may curve through
the crystal.

\engterm{Dislocation motion}. An edge dislocation under shear stress
moves laterally by successive bond rearrangements: the half-plane
shifts one lattice spacing, then another, and another, sweeping
through the crystal until it exits at a free surface. The stress
required for the motion is the \engterm{Peierls-Nabarro stress},
much smaller than the theoretical shear strength because at any
moment only a small number of bonds at the dislocation core are
rearranging. A screw dislocation moves perpendicular to its Burgers
vector by the same general mechanism but with cross-slip available:
a screw dislocation can switch from one slip plane to another that
contains its Burgers vector, allowing it to move around obstacles.

\engterm{Plastic strain and dislocation density}. The plastic shear
strain produced by the motion of a single dislocation a distance
$\Delta x$ is $\Delta \gamma = b \Delta x / V$ for a crystal of
volume $V$ and lattice spacing $b$. For a population of $N$
dislocations each moving the average distance $\bar{\Delta x}$,
\[
\gamma = \rho b \bar{\Delta x},
\]
where $\rho = N/V$ is the dislocation density. Annealed metals have
$\rho \sim 10^{10}\,\text{m}^{-2}$ (one dislocation per
square micrometre); heavily cold-worked metals have $\rho \sim 10^{14}
\,\text{m}^{-2}$. The increased density makes further plastic
deformation harder because the dislocations interact with each other,
the phenomenon known as \engterm{strain hardening} or
\engterm{work hardening}.

\engterm{Stress to move a dislocation against obstacles}. The
\engterm{flow stress} $\tau$ required to push a dislocation past
obstacles scales as
\[
\tau = \alpha G b \sqrt{\rho},
\]
the Taylor relation, where $\alpha \sim 0.2$ to $0.5$ is a numerical
factor and $G$ is the shear modulus. The relation predicts that the
flow stress increases as the square root of the dislocation density,
consistent with the strain-hardening behaviour observed in tensile
tests of cold-worked metals.

\engterm{Schmid's law and resolved shear stress}. A single crystal
under uniaxial tension at stress $\sigma$ feels a resolved shear stress
on a chosen slip system,
\[
\tau = \sigma \cos\phi \cos\lambda,
\]
where $\phi$ is the angle between the tensile axis and the slip-plane
normal, and $\lambda$ is the angle between the tensile axis and the
slip direction. The product $\cos\phi \cos\lambda$ is the
\engterm{Schmid factor}; slip begins on the slip system with the
highest Schmid factor when the resolved shear stress equals the
critical resolved shear stress of the lattice. For a polycrystalline
metal under uniaxial tension, the Schmid factor averaged over all
crystallographic orientations is the \engterm{Taylor factor} $M \approx
3.06$ for FCC and $M \approx 2.75$ for BCC; the macroscopic yield
stress is $M$ times the single-crystal critical resolved shear stress.

\section{Strengthening mechanisms}\pathtag{core}

Engineering metals are strengthened by four principal mechanisms,
each of which raises the flow stress by impeding dislocation motion.
The mechanisms are largely additive (with caveats) and the engineer
combines them to reach the design strength.

\engterm{Solid-solution strengthening}. Substitutional or interstitial
solute atoms produce local lattice distortions that interact
elastically with passing dislocations. The flow-stress contribution
scales with solute concentration and with the size or modulus
mismatch between solute and solvent. For a substitutional solute with
size misfit $\varepsilon = \Delta a/a$ (the fractional lattice
parameter change per atomic fraction of solute), the strengthening
contribution is approximately $\Delta \tau \sim G \varepsilon^{3/2}
c^{1/2}$, where $c$ is the solute concentration. Carbon and nitrogen
in iron (interstitial solutes) produce particularly strong
solid-solution strengthening because the interstitial site is small
and the misfit is large; copper-zinc brasses have $\sim 100\,\text{MPa}$
of solid-solution strengthening from the zinc.

\engterm{Strain hardening (work hardening)}. The flow stress increases
with prior plastic deformation, by the Taylor relation $\tau \propto
\sqrt{\rho}$ as discussed above. The relation between flow stress and
strain in a cold-worked metal is empirically captured by the
\engterm{Hollomon equation}
\[
\sigma = K \varepsilon^n,
\]
with $K$ the strength coefficient and $n$ the strain-hardening
exponent. For low-carbon steel, $n \approx 0.2$; for $70\text{-}30$
brass, $n \approx 0.5$; for aluminium 1100, $n \approx 0.2$.
High-$n$ materials show large hardening at small strains and are
preferred for sheet-forming applications (deep drawing of beverage
cans, automotive body panels) because the diffuse necking that
$n \approx 0.5$ permits gives uniform thickness rather than
concentrated thinning.

\engterm{Grain-size strengthening (Hall-Petch)}. Grain boundaries are
obstacles to dislocation motion: a dislocation moving in one grain
piles up against the boundary and cannot pass directly into the
neighbouring grain (which has a different crystallographic
orientation). The stress required for slip to propagate from one
grain to the next scales as the inverse square root of the grain
size,
\[
\sigma_y = \sigma_0 + k d^{-1/2},
\]
the \engterm{Hall-Petch relation}, with $d$ the average grain
diameter, $\sigma_0$ the intrinsic single-crystal yield stress, and
$k$ a material-specific constant ($\sim 0.7\,\text{MPa}\cdot\text{mm}^{1/2}$
for low-carbon steel, $\sim 0.4\,\text{MPa}\cdot\text{mm}^{1/2}$ for
aluminium). Refining grain size from $100\,\si{\micro\meter}$ to
$10\,\si{\micro\meter}$ approximately doubles the yield strength of
low-carbon steel. The Hall-Petch relation breaks down at very small
grain sizes ($< 100\,\text{nm}$) because grain-boundary sliding
becomes the dominant deformation mechanism; the inverse Hall-Petch
regime is the subject of an active research literature in
nanocrystalline metals.

\engterm{Precipitation strengthening (age hardening)}. A second-phase
precipitate produced by controlled heat treatment of a supersaturated
solid solution impedes dislocation motion. The mechanism depends on
the precipitate's size and coherence: small coherent precipitates
are cut by passing dislocations, contributing $\Delta \tau \propto
\sqrt{f r}$ (with $f$ the volume fraction and $r$ the precipitate
radius); large incoherent precipitates are bypassed by dislocation
looping (the Orowan mechanism), contributing $\Delta \tau \propto
\sqrt{f}/r$. The two contributions cross over at a critical
precipitate radius (typically $5$ to $50\,\text{nm}$), and the
maximum strengthening occurs near the crossover. Age hardening is
the strengthening mechanism of the duralumin family (Al-Cu, Al-Cu-Mg
alloys), of the nickel-based superalloys (Ni-Al-Ti $\gamma'$
precipitates), and of the maraging steels (intermetallic precipitates
in a low-carbon iron-nickel matrix). The process was discovered
empirically by Alfred Wilm in 1906; the precipitation mechanism was
worked out in the 1930s and 1940s.

\begin{principle}[Strengthening is dislocation obstruction]
A metal's yield strength equals the friction stress of pure single
crystal plus the contributions from solid-solution, strain-hardening,
grain-boundary, and precipitation obstacles. The four contributions
add (approximately, in the dilute-obstacle limit; in quadrature when
obstacle distributions are comparable; with more involved combinations
when one mechanism dominates). The engineer who specifies a high-
strength alloy combines the four mechanisms by choice of composition,
processing, and heat treatment, with each choice committing the
component to a specific microstructural path.
\end{principle}

\section{Steels and cast irons}\pathtag{standard}

Steel is the dominant structural material of modern infrastructure.
Annual global production is approximately $1.9\,\text{Gt}$ as of
2025, an order of magnitude larger than the next-most-produced
metal (aluminium, $\sim 70\,\text{Mt}$). The dominance comes from
the combination of high specific strength, ease of fabrication,
recyclability, and low cost per kilogram ($\sim \$0.6$ to $\$1.0/
\text{kg}$ for hot-rolled commodity grades in 2025).

\engterm{Plain-carbon steels}. Iron-carbon alloys with carbon in
the $0.05$ to $1.5\,\%$ range and minimal other alloying. The
AISI/SAE designation system uses four digits: the first two for the
alloy class, the last two for the carbon content in hundredths of a
percent. AISI 1020 ($0.20\,\%$ C) is structural steel for low-stress
applications; 1045 ($0.45\,\%$ C) is medium-carbon steel for shafts
and gears; 1090 ($0.90\,\%$ C) is high-carbon steel for springs and
cutting tools. The yield strength in the as-rolled condition is
$\sim 250$ to $700\,\text{MPa}$, increasing with carbon content;
quench-and-temper heat treatment raises the yield strength further.

\engterm{Low-alloy steels}. Plain-carbon compositions with deliberate
additions of manganese, chromium, molybdenum, nickel, or vanadium
to improve hardenability and elevated-temperature strength. The AISI
4xxx series contains molybdenum (4140 is $0.4\,\%$ C plus Cr and Mo);
8xxx contains nickel-chromium-molybdenum. Low-alloy steels are the
working choice for high-strength fasteners (8740 for grade-8 bolts),
shafts (4140 for general-purpose), and pressure vessels (specific
chromium-molybdenum compositions for elevated-temperature service).

\engterm{Stainless steels}. Iron-chromium alloys with chromium content
exceeding $10.5\,\%$ form a passive chromium-oxide film that confers
corrosion resistance. The major families:

\begin{itemize}[leftmargin=*]
\item \engterm{Austenitic stainless} (304, 316, 321 series): FCC at
  room temperature, stabilised by nickel (typically $8$ to $20\,\%$).
  Not heat-treatable by quench-and-temper; strengthened by cold
  working or by nitrogen alloying. Excellent corrosion resistance,
  formability, and weldability. The dominant stainless family for
  food-contact and chemical-process applications. 316 has molybdenum
  ($\sim 2\,\%$) added for resistance to chloride pitting.
\item \engterm{Ferritic stainless} (430, 444): BCC at room temperature,
  no nickel. Magnetic, less ductile than austenitic, less expensive.
  Common for automotive trim, kitchen appliances.
\item \engterm{Martensitic stainless} (410, 420, 440): can be hardened
  by quench-and-temper. Higher carbon than ferritic. Common for
  cutlery, surgical instruments, and high-strength corrosion-
  resistant components.
\item \engterm{Duplex stainless} (2205, 2507): roughly equal volumes
  of austenitic and ferritic phases. Higher strength than either
  parent family. Common for offshore and marine structural
  applications.
\item \engterm{Precipitation-hardening stainless} (17-4 PH, 15-5 PH):
  martensitic matrix with copper or aluminium precipitates from a
  controlled ageing treatment. High strength combined with corrosion
  resistance. Common for aerospace and high-performance applications.
\end{itemize}

\engterm{Tool steels}. Highly alloyed steels designed for cutting,
forming, or shaping other materials. The AISI tool-steel classification
uses letter prefixes: W (water-hardening, plain-carbon), O (oil-
hardening), A (air-hardening), D (cold-work die), H (hot-work),
S (shock-resisting), M (molybdenum-based high-speed), T (tungsten-
based high-speed). Tool steels achieve their hardness through high
carbon content and through complex carbides (chromium, vanadium,
tungsten, molybdenum carbides) that resist softening at elevated
temperatures.

\engterm{Cast irons}. Iron-carbon alloys with carbon above the
maximum austenite solubility ($\sim 2.14\,\%$), typically $2.5$ to
$4.0\,\%$ C, plus $1$ to $3\,\%$ silicon. The carbon precipitates as
graphite (in grey, ductile, and malleable irons) or as cementite (in
white iron). The graphite morphology distinguishes the working types:

\begin{itemize}[leftmargin=*]
\item \engterm{Grey iron}: graphite as flakes embedded in a ferrite
  or pearlite matrix. The flakes are stress concentrators that
  reduce tensile strength but provide excellent vibration damping
  and high compressive strength. Used for engine blocks, machine
  bases, and components in compression service.
\item \engterm{Ductile iron} (nodular cast iron): graphite as
  spheres, produced by magnesium or cerium additions that suppress
  flake growth. The spheroidal graphite removes the stress
  concentrations of grey iron, giving tensile strength and ductility
  comparable to cast steel. Used for pipes, gears, valve bodies, and
  general-purpose castings.
\item \engterm{White iron}: no graphite; carbon present as cementite.
  Very hard ($\sim 600\,\text{HB}$) and very brittle. Used for wear-
  resistant surfaces, often as a chilled outer layer on a grey-iron
  casting.
\item \engterm{Malleable iron}: white iron heat-treated to convert
  the cementite to graphite clusters. More ductile than white iron,
  competing with ductile iron in some applications. Largely replaced
  by ductile iron since the 1950s.
\end{itemize}

\engterm{Production volumes and trends}. The bulk of global steel
production is plain-carbon and low-alloy steel; stainless and tool
steels combined account for $\sim 3\,\%$ of production by mass. Cast
irons account for $\sim 5\,\%$. The post-2000 trend is toward
high-strength low-alloy steels (HSLA, with yield strengths in the
$400$ to $700\,\text{MPa}$ range) that allow lighter sections in
automotive, structural, and pipeline applications, and toward
advanced high-strength steels (AHSS, with yield strengths exceeding
$1\,\text{GPa}$ and engineered transformation behaviour) for
automotive structural components.

\section{Aluminium alloys}\pathtag{standard}

Aluminium is the second-most-produced structural metal globally
($\sim 70\,\text{Mt}/\text{yr}$ as of 2025). Pure aluminium is soft
and ductile; engineering aluminium alloys derive their strength from
solid-solution, strain-hardening, and precipitation mechanisms,
typically in combination.

\engterm{Wrought aluminium designation}. The Aluminum Association's
four-digit system: the first digit identifies the principal alloying
element, the remaining three identify the specific alloy.

\begin{itemize}[leftmargin=*]
\item 1xxx (pure aluminium, $\geq 99\,\%$ Al): conductor grade, soft
  and ductile, low strength. Used for electrical bus bar and
  chemical-process tanks.
\item 2xxx (Al-Cu): high strength, heat-treatable, moderate corrosion
  resistance. Duralumin family. Used for aerospace structural
  components (2014, 2024, 2219).
\item 3xxx (Al-Mn): moderate strength, formable, used for beverage
  cans (3004) and architectural sheet.
\item 4xxx (Al-Si): used for welding and brazing filler metals,
  occasionally for casting.
\item 5xxx (Al-Mg): non-heat-treatable, strain-hardened. Excellent
  corrosion resistance, especially in marine environments (5052,
  5083, 5754). Used for ship hulls, pressure vessels, and automotive
  body panels.
\item 6xxx (Al-Mg-Si): heat-treatable, moderate strength, good
  formability and weldability. The dominant extrusion alloy
  (6061, 6063). Used for structural extrusions, architectural
  components, and automotive frame components.
\item 7xxx (Al-Zn-Mg, occasionally Al-Zn-Mg-Cu): highest-strength
  aluminium alloys (7075, 7050, 7475). Used for aircraft structural
  components and high-performance bicycle frames.
\item 8xxx (Al-Li and others): specialty alloys with lithium
  additions for reduced density and higher modulus. Used for advanced
  aerospace components.
\end{itemize}

\engterm{Temper designations}. The Aluminum Association's letter-
number temper system follows the alloy designation. F (as fabricated),
O (annealed), H (strain-hardened), T (heat-treated). The H is
followed by two digits indicating the strain-hardening type and
degree. The T is followed by a digit indicating the heat-treatment
sequence (T6 is solution-treated and artificially aged, the most
common temper for 6xxx and 7xxx alloys).

\engterm{Age hardening}. Wilm's 1906 observation that an
aluminium-copper alloy hardens spontaneously at room temperature
after solution treatment is the founding example of precipitation
strengthening. The modern aluminium-copper-magnesium alloy 2024-T351
(solution-treated, stretched, and naturally aged) reaches yield
strength $\sim 325\,\text{MPa}$ from a base aluminium of $\sim 35\,
\text{MPa}$. The 7075-T6 alloy (solution-treated, artificially aged
at $120\,\degC$ for $24\,\text{h}$) reaches $\sim 500\,\text{MPa}$,
making it the standard high-strength wrought aluminium for aircraft.

\engterm{Cast aluminium}. The 3xx.x cast designation (A356, A357,
A380) for aluminium-silicon-magnesium alloys is the dominant casting
family, accounting for the bulk of die-cast automotive components.
The silicon (typically $5$ to $20\,\%$) lowers the melting point and
improves castability; the magnesium (typically $0.3$ to $0.7\,\%$)
permits heat treatment. Aluminium die castings dominate the
production of internal combustion engine blocks, transmission
housings, structural brackets, and electronic enclosures.

\section{Copper, titanium, magnesium, nickel-based superalloys}\pathtag{standard}

\subsection{Copper alloys}

Pure copper has the highest electrical conductivity of any commercial
material at room temperature ($5.96 \times 10^7\,\text{S}/\text{m}$,
or $58\,\text{MS}/\text{m}$; silver is slightly higher but too
expensive for bulk applications). Copper is also the second-highest
thermal conductor. Engineering use is split among electrical
conductors, plumbing tube, heat exchanger tube, electronic
substrates, decorative architecture, and the major copper alloys.

\engterm{Brasses} (Cu-Zn). The most-used copper alloy family.
Cartridge brass (70-30) is the deep-drawing alloy for ammunition
cases, plumbing fittings, and decorative hardware. Yellow brass
(65-35) is the architectural and decorative standard. Naval brass
(60-40 with small tin addition) is used in marine fittings.

\engterm{Bronzes} (Cu-Sn, Cu-Al, Cu-Be, Cu-Si). The tin bronzes are
the historical engineering material of antiquity and remain in use
for bearings and bushings (Cu-Sn-Pb), bells (Cu-Sn high tin), and
ornamental castings. Aluminium bronzes have higher strength than
tin bronzes and resist sea-water corrosion. Beryllium copper (Cu with
$\sim 2\,\%$ Be) is the highest-strength copper alloy ($\sim 1300\,
\text{MPa}$ yield in age-hardened condition), used for non-sparking
tools, electrical springs, and specialty connectors; the beryllium
content is a health hazard during machining and welding.

\engterm{Copper-nickel} (cupronickel, monel-class). Used for marine
heat exchangers, condenser tubes, and coinage (75-25 cupronickel is
the US nickel five-cent coin). Excellent corrosion resistance in
sea water.

\subsection{Titanium alloys}

Titanium has high specific strength (yield strength over density),
excellent corrosion resistance (a passive titanium-oxide film), and
biocompatibility. The disadvantages are high cost ($\sim \$10$ to
$\$50/\text{kg}$ for mill product as of 2025), poor machinability
relative to aluminium or steel, and the difficulty of casting and
welding. Engineering titanium alloys are classified by phase
behaviour:

\begin{itemize}[leftmargin=*]
\item \engterm{$\alpha$ alloys} (commercially pure Ti, Ti-5Al-2.5Sn):
  HCP at all engineering temperatures. Not heat-treatable. Good
  weldability, used for cryogenic applications and chemical-process
  service.
\item \engterm{$\alpha+\beta$ alloys} (Ti-6Al-4V is the workhorse,
  accounting for $\sim 50\,\%$ of titanium production): two-phase
  microstructure, heat-treatable, used for aircraft structural
  components, gas turbine compressor blades, and biomedical
  implants.
\item \engterm{$\beta$ alloys} (Ti-3Al-8V-6Cr-4Mo-4Zr): BCC at room
  temperature, heat-treatable to higher strength than $\alpha+\beta$.
  Used for high-strength fastener and spring applications.
\end{itemize}

The aerospace industry consumes the majority of mill-product titanium
($\sim 60\,\%$ as of 2025); chemical-process service, marine
applications, and biomedical implants account for the rest.

\subsection{Magnesium alloys}

Magnesium has the lowest density of any structural metal ($1.74\,
\text{g}/\text{cm}^3$, two-thirds that of aluminium). HCP crystal
structure produces the anisotropy and limited room-temperature
ductility discussed in Section~1 of this chapter. Engineering
applications are dominated by die-cast automotive components
(magnesium-aluminium-zinc alloys, AZ91 being the dominant grade), by
laptop and camera bodies, and by aerospace components where the
density advantage compensates for the higher cost and the
poorer ductility. Magnesium burns readily in air at elevated
temperature; magnesium machining swarf and dust are fire hazards, and
the molten alloy must be protected by a covering gas during casting.

\subsection{Nickel-based superalloys}

The high-temperature working alloys of gas turbines, jet engines, and
rocket engines. The matrix is FCC nickel, alloyed with chromium for
oxidation resistance, cobalt for solid-solution strengthening, and
aluminium-titanium for the precipitation of the strengthening
$\gamma'$ phase Ni$_3$(Al,Ti). The $\gamma'$ precipitate is coherent
with the $\gamma$ matrix and has the anomalous property that its
yield strength increases with temperature up to $\sim 800\,\degC$;
the macroscopic alloy strength is preserved or even increased over
much of the engine's operating temperature range. Refractory metals
(tungsten, rhenium, tantalum) and reactive elements (hafnium,
yttrium) provide further strengthening and oxidation resistance.

The major superalloy families:

\begin{itemize}[leftmargin=*]
\item Inconel 718 (Ni-Fe-Cr-Nb): the dominant superalloy by volume,
  used for gas-turbine disks, casings, and structural components up
  to $\sim 650\,\degC$.
\item Inconel 738, Mar-M-247, CMSX series: single-crystal castings
  for turbine blades, used up to $\sim 1100\,\degC$ with thermal
  barrier coatings (Volume IV Ch~6 develops the coating systems).
\item Hastelloy alloys (Ni-Cr-Mo): corrosion-resistant rather than
  high-temperature; used for chemical-process equipment in aggressive
  environments.
\end{itemize}

Production volumes of nickel-based superalloys are small ($\sim 1
\,\text{Mt}/\text{yr}$ globally) but the alloys' specific
applications (high-pressure turbine blades, hot-section components,
combustor liners) are mission-critical.

\section{Welding and joining metals}\pathtag{standard}

A welded joint produces a continuous metallurgical bond between two
metal components by local melting and solidification. The joint's
strength can equal or exceed that of the parent metal in well-
designed and well-executed welds; the joint's defects can produce
crack-initiating sites that fail at stresses far below the parent
metal's design capacity.

\engterm{Welding processes}. The dominant industrial welding methods:

\begin{itemize}[leftmargin=*]
\item \engterm{Shielded metal arc welding (SMAW)}: a consumable
  flux-coated electrode struck against the work establishes an
  electric arc. The arc melts the electrode and the work; the flux
  shields the molten metal from atmospheric oxygen. Common manual
  process for field welding of structural steel.
\item \engterm{Gas metal arc welding (GMAW, MIG)}: a continuously fed
  wire electrode with an external shielding gas (argon, carbon
  dioxide, or mixture). Higher productivity than SMAW; used for
  production welding of steel and aluminium.
\item \engterm{Gas tungsten arc welding (GTAW, TIG)}: a non-consumable
  tungsten electrode with separate filler-rod addition and inert
  shielding gas. Slower than GMAW but produces high-quality joints in
  thin sections and reactive metals (titanium, aluminium).
\item \engterm{Submerged arc welding (SAW)}: a granular flux blanket
  covers the arc, providing shielding and slag formation. Used for
  high-deposition-rate welding of thick steel sections (pipelines,
  pressure vessels, structural beams).
\item \engterm{Resistance welding} (spot, seam, butt): the work pieces
  themselves carry the welding current; resistance heating at the
  interface produces the weld. Common for sheet-metal automotive
  assembly.
\item \engterm{Laser and electron-beam welding}: high-energy-density
  processes producing narrow, deep welds with small heat-affected
  zones. Used for precision applications.
\end{itemize}

\engterm{The heat-affected zone}. A region adjacent to the fusion
zone experiences temperatures below melting but high enough to alter
the microstructure of the parent metal. For carbon and low-alloy
steels, the heat-affected zone (HAZ) shows a gradient of
microstructure from coarse-grained austenite-transformed (adjacent
to the weld) through fine-grained recrystallised (intermediate) to
spheroidised pearlite or unaffected base metal (at the outer
boundary). The HAZ's mechanical properties differ from both the
weld metal and the base metal, and the joint's failure most often
initiates at the HAZ boundary where toughness is lowest.

\engterm{Filler metals}. Most welding processes use a filler metal
(electrode wire or filler rod) with composition similar to but not
identical to the base metal. For carbon steels, the filler is
typically a low-alloy steel with controlled silicon, manganese, and
nickel to manage the weld's deoxidation and solidification. For
stainless steels, the filler is matched to the base alloy with
adjustments for dilution and ferrite-content control. For aluminium
alloys, fillers are selected from the 4xxx series for many parent
alloys, with attention to crack sensitivity and corrosion
compatibility.

\section{Failure: weld defects, hydrogen embrittlement, stress-corrosion cracking}\pathtag{core}

Three failure modes are specific to engineering metals and the
joining processes used to assemble them. Each has produced consequential
service failures and each has a well-established mitigation chain that
the design and manufacturing engineer must follow.

\engterm{Weld defects}. A welded joint can contain any of the
following defect types, each visible in cross-section and detectable
by non-destructive evaluation:

\begin{itemize}[leftmargin=*]
\item \engterm{Lack of fusion}: the filler metal does not bond to
  the parent metal along part of the joint. Produced by insufficient
  heat input, insufficient cleaning, or wrong torch angle. Acts as
  a pre-existing crack under fatigue loading.
\item \engterm{Lack of penetration}: the filler metal does not reach
  the root of the joint, leaving an unwelded gap. Common in
  manual welding of thick sections.
\item \engterm{Porosity}: gas bubbles trapped in the solidifying
  weld. Sources include moisture in the filler, contamination on
  the work surface, and atmospheric infiltration through inadequate
  shielding gas.
\item \engterm{Slag inclusions}: solidified flux particles embedded
  in the weld. Common in multi-pass SMAW welds where inter-pass
  cleaning was incomplete.
\item \engterm{Cracks}: solidification cracks (hot cracks) in the
  fusion zone, and cold cracks in the HAZ. Hot cracks form at the
  end of solidification when residual stress exceeds the strength of
  the partially solidified metal. Cold cracks form during cool-down
  of high-strength steels and are typically associated with hydrogen.
\item \engterm{Undercut}: a groove melted into the parent metal
  adjacent to the weld toe and not filled with filler metal. Acts as
  a stress concentrator.
\end{itemize}

Weld inspection follows standardised codes (ASME Boiler and Pressure
Vessel Code Section IX for pressure vessels; AWS D1.1 for structural
steel) that specify visual inspection, dye-penetrant or
magnetic-particle inspection of surfaces, radiographic or
ultrasonic inspection of internal volumes, and acceptance criteria
for each defect type.

\engterm{Hydrogen embrittlement}. High-strength steels (yield
strength $\gtrsim 1000\,\text{MPa}$) absorb atomic hydrogen from
service environments (corrosion in aqueous environments,
electroplating, pickling, welding with moisture in the flux). The
absorbed hydrogen diffuses through the steel and segregates to
dislocation cores, grain boundaries, and the tips of stress
concentrators. At sufficient concentration the hydrogen lowers the
local cohesive energy of the lattice and the steel fractures by
intergranular or transgranular cleavage at stresses well below the
nominal yield strength. The mechanism (hydrogen-enhanced decohesion,
or hydrogen-enhanced localised plasticity, depending on the alloy
and conditions) is the subject of an active literature; the
engineering consequence is consistent across mechanisms.

\engterm{Failure pattern}. A typical hydrogen-embrittlement failure
follows a delayed-rupture pattern: a high-strength fastener or
spring is installed and torqued to specification, operates in
service for hours to weeks, and then fails by intergranular brittle
fracture without warning. The fracture surface shows fish-eye
markings around inclusions where hydrogen accumulated. The failure
is reproducible: replacement parts from the same heat treatment
fail in the same time-to-rupture range under the same conditions.
The standard mitigations:

\begin{itemize}[leftmargin=*]
\item Avoid hydrogen-introducing process steps where possible. Cadmium
  plating, the traditional corrosion-resistant coating for high-
  strength fasteners, is hydrogen-introducing and is being replaced
  by mechanically applied zinc-nickel coatings or by sealed
  hot-dipped aluminium coatings.
\item If hydrogen-introducing steps are unavoidable, apply a baking
  treatment (typically $190$ to $220\,\degC$ for $4$ to $24\,
  \text{h}$) immediately after the step to drive hydrogen out
  before the lattice equilibrates with the trapped hydrogen.
\item Limit the strength of fasteners used in environments that
  could re-introduce hydrogen in service (cathodic-protected marine
  structures, hydrogen-bearing process equipment). Stainless steels
  in the austenitic family resist hydrogen embrittlement because
  hydrogen has very low diffusivity in FCC lattices; austenitic
  stainless is the standard for fasteners in hydrogen service.
\end{itemize}

\engterm{Stress-corrosion cracking (SCC)}. Certain combinations of
alloy and environment produce cracks that propagate at applied
stresses far below the alloy's yield strength when both stress and
corrosive environment are present simultaneously. The cracks are
typically branched and transgranular or intergranular depending on
the alloy-environment combination. Each susceptible combination has
a specific environment that triggers it:

\begin{itemize}[leftmargin=*]
\item Austenitic stainless steel + chloride solution at $\gtrsim
  60\,\degC$: transgranular chloride SCC. The mitigation is to avoid
  chlorides in service, to use a more SCC-resistant alloy (duplex
  stainless, nickel-based alloys), or to design for compressive
  surface residual stresses.
\item Brass + ammonia or moisture: ``season cracking,'' the classical
  SCC of the brass era. Mitigation includes stress relief
  annealing and avoiding ammoniacal environments.
\item Carbon steel + concentrated caustic solution at elevated
  temperature: ``caustic cracking.'' Mitigated by post-weld stress
  relief and by limiting service temperatures.
\item High-strength aluminium (7xxx series) + chloride solution:
  intergranular SCC. Mitigated by over-ageing treatments (T73, T76
  tempers) that sacrifice some yield strength for SCC resistance.
\end{itemize}

\engterm{Detection and prevention}. SCC failures often initiate from
microscopic surface flaws that are not detected by routine
inspection. The standard prevention chain combines materials
selection (chose an alloy with no known SCC vulnerability in the
service environment), stress management (specify allowable stress
below the threshold stress intensity for crack propagation,
$K_{\text{ISCC}}$), and environment control (limit chloride, ammonia,
caustic, or other susceptibility-causing species). The Boeing 707
example of SCC in 7079-T6 wing-spar fittings, identified in service
in the early 1960s and producing a fleet-wide retrofit, was the
canonical demonstration of the failure mode and the impetus for the
development of the more SCC-resistant 7050 and 7475 alloys for
aerospace structural use.

\engterm{Related failure: weld-zone SCC}. The combination of welding
(which introduces residual stresses and an altered microstructure)
and a susceptibility-causing environment produces a high-risk
location. Pipeline weld-zone SCC has been a major maintenance
concern for natural-gas transmission lines; the mitigation chain
includes post-weld heat treatment, coating, cathodic protection, and
in-service inspection.

\begin{principle}[Three failure pairings]
The three failure modes of engineering metals (weld defects,
hydrogen embrittlement, stress-corrosion cracking) all combine a
microstructural vulnerability with an environmental trigger. A
defect-free weld in a hydrogen-free atmosphere in an SCC-resistant
alloy environment has none of the three. The engineering response
combines materials selection, manufacturing-process control, and
service-environment management; the three lines of defence are
independent, and dropping any one of them initiates the corresponding
failure mode.
\end{principle}

\begin{estimation}
\textbf{Question.} Estimate the yield strength of pure FCC aluminium
($\sigma_y \approx 35\,\text{MPa}$), then estimate the yield strength
gain from refining grain size from $100\,\si{\micro\meter}$ to
$10\,\si{\micro\meter}$, and from $10\,\si{\micro\meter}$ to $1\,
\si{\micro\meter}$.

\smallskip
\textit{Estimate, before reading on.}

\smallskip
Hall-Petch: $\sigma_y = \sigma_0 + k d^{-1/2}$. For aluminium, $\sigma_0
\approx 20\,\text{MPa}$ (single-crystal friction stress) and $k \approx
0.07\,\text{MPa}\cdot\text{m}^{1/2}$ (rounded value for pure aluminium
\cite[ch.~6]{text:dieter-mechanical-metallurgy}).

For $d = 100\,\si{\micro\meter} = 10^{-4}\,\text{m}$: $d^{-1/2} = 100\,
\text{m}^{-1/2}$, so $\sigma_y = 20 + 0.07 \times 100 = 27\,\text{MPa}$.
For $d = 10\,\si{\micro\meter} = 10^{-5}\,\text{m}$: $d^{-1/2} = 316\,
\text{m}^{-1/2}$, so $\sigma_y = 20 + 0.07 \times 316 = 42\,\text{MPa}$.
For $d = 1\,\si{\micro\meter} = 10^{-6}\,\text{m}$: $d^{-1/2} = 1000\,
\text{m}^{-1/2}$, so $\sigma_y = 20 + 0.07 \times 1000 = 90\,\text{MPa}$.

The grain refinement from $100$ to $10\,\si{\micro\meter}$ adds
$\sim 15\,\text{MPa}$; from $10$ to $1\,\si{\micro\meter}$ adds
another $48\,\text{MPa}$. Refining from $100$ to $1\,\si{\micro\meter}$
roughly triples the yield strength.

\smallskip
\textit{Postmortem.} The estimate is internally consistent and matches
the empirical Hall-Petch slope for pure aluminium to within published
ranges. The Hall-Petch relation begins to fail below $\sim 100\,\text{nm}$
because grain-boundary sliding becomes the dominant deformation mode
(the inverse Hall-Petch regime). Conventional rolling and recrystallisation
can produce aluminium grain sizes down to $\sim 10\,\si{\micro\meter}$;
severe plastic deformation processes (equal-channel angular pressing,
high-pressure torsion) can produce ultrafine grain sizes below $1\,
\si{\micro\meter}$, with the corresponding yield-strength gains. The
calculation is the working tool for estimating how much strengthening
to expect from a chosen process route.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Identify an unknown metal sample]
The reader is given (or selects from household items) an unknown
metal sample and identifies its alloy class by a sequence of
household tests: density measurement, hardness comparison, magnetic
attraction, surface colour and oxidation, and (if a multimeter is
available) electrical conductivity.

\textbf{Track}. Household identification.
\textbf{Hazard class}: low. No furnaces, no chemical reagents.

\textbf{Procedure}.

\begin{enumerate}[leftmargin=*]
\item Select an unknown sample (a fragment of plumbing, a discarded
  fastener, a piece of cookware, a coin) and measure its mass on a
  kitchen scale to $0.1\,\text{g}$ precision.
\item Measure its volume by water displacement: place the sample in
  a graduated container, note the water level, immerse the sample
  fully, note the new water level, and subtract. Compute the
  density $\rho = m/V$.
\item Test magnetism with a refrigerator magnet. Ferromagnetic
  attraction suggests iron, steel, nickel, or cobalt; no attraction
  is consistent with aluminium, copper, brass, stainless steel
  (austenitic family), or zinc.
\item Compare surface hardness by trying to scratch with progressively
  harder objects: fingernail (Mohs $\sim 2.5$), copper coin ($\sim
  3$), glass plate ($\sim 5.5$), steel file ($\sim 6.5$). Note the
  hardest object that fails to scratch the sample and the softest
  that succeeds.
\item If a multimeter is available, measure the electrical resistance
  of a short section of the sample and compute its resistivity from
  the dimensions. Compare to handbook values.
\item Observe the sample's surface colour (silvery, grey, red-brown,
  yellow), the colour of any oxide film, and the qualitative
  surface finish.
\item Compile the measurements into a one-page decision table and
  identify the alloy class consistent with all measurements.
\end{enumerate}

\textbf{Analysis}.

\begin{itemize}[leftmargin=*]
\item Compare measured density to handbook values: aluminium $2.70\,
  \text{g}/\text{cm}^3$, magnesium $1.74$, titanium $4.51$, iron
  $7.87$, copper $8.96$, brass $8.4$ to $8.7$, stainless steel
  $7.8$ to $8.0$, lead $11.34$.
\item Identify the alloy class whose properties are most consistent
  with the measurements. Identify any property that does not fit and
  propose a refinement (e.g. heat-treatment temper, alloy content)
  that resolves the discrepancy.
\item For one of the samples, predict one property not measured
  (yield strength, thermal conductivity, melting point) from the
  identified alloy class.
\end{itemize}

\textbf{Deliverable}. A two- to four-page report with the
measurement table, the identification, and the predicted property.

\textbf{Time}. Two hours of bench work, one to two hours of write-up.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Compute the planar density of atoms on the $\{111\}$ slip planes of
FCC copper ($a = 3.61\,\si{\angstrom}$) and on the $\{110\}$ slip
planes of BCC iron ($a = 2.87\,\si{\angstrom}$). Compare and identify
which planes are more closely packed.
\end{exercise}

\begin{exercise}
For FCC aluminium ($a = 4.05\,\si{\angstrom}$), compute the Burgers
vector magnitude $b = a/\sqrt{2}$ for the $\langle 110 \rangle$ slip
direction. Compute the resolved shear stress required to move a
dislocation against an obstacle density $\rho = 10^{12}\,\text{m}^{-2}$
using the Taylor relation with $\alpha = 0.3$ and shear modulus
$G = 26\,\text{GPa}$.
\end{exercise}

\begin{exercise}
A low-carbon steel has yield strength $\sigma_y = 200\,\text{MPa}$ at
grain size $50\,\si{\micro\meter}$ and $\sigma_y = 350\,\text{MPa}$ at
grain size $5\,\si{\micro\meter}$. Determine the Hall-Petch parameters
$\sigma_0$ and $k$.
\end{exercise}

\begin{exercise}
A copper rod is cold-rolled to $50\,\%$ reduction in cross-section.
Estimate the resulting yield-strength increase using the Hollomon
equation with $K = 530\,\text{MPa}$ and $n = 0.33$. State the
assumptions.
\end{exercise}

\begin{exercise}
For a 7075-T6 aluminium alloy with yield strength $500\,\text{MPa}$
and density $2.81\,\text{g}/\text{cm}^3$, compute the specific
strength ($\sigma_y/\rho$). Compare to AISI 4340 quenched-and-tempered
steel ($\sigma_y = 1000\,\text{MPa}$, $\rho = 7.85\,\text{g}/\text{cm}^3$)
and to Ti-6Al-4V ($\sigma_y = 880\,\text{MPa}$, $\rho = 4.43\,\text{g}/
\text{cm}^3$).
\end{exercise}

\begin{exercise}
Compute the Schmid factor for a uniaxial tensile test on a single-
crystal FCC metal with the tensile axis along $[110]$ and slip on the
$(111)[\bar{1}01]$ system.
\end{exercise}

\begin{exercise}
A nickel-based superalloy operates at $1000\,\degC$ in a turbine
hot section. The yield strength at this temperature is $700\,\text{MPa}$,
and the alloy's $0.1\,\%$ creep strain criterion limits stress to
$300\,\text{MPa}$ for a $10\,000\,\text{h}$ service life. Identify
which criterion governs the allowable stress and the engineering
margin remaining.
\end{exercise}

\begin{exercise}
A weld in low-carbon steel produces a heat-affected zone of width
$5\,\text{mm}$ with grain size coarsened from $30\,\si{\micro\meter}$
in the base metal to $100\,\si{\micro\meter}$ in the HAZ. Estimate
the HAZ yield strength relative to the base metal using Hall-Petch
with $\sigma_0 = 100\,\text{MPa}$ and $k = 0.7\,\text{MPa}\cdot\text{mm}^{1/2}$.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive Schmid's law $\tau = \sigma \cos\phi \cos\lambda$ for the
resolved shear stress on a slip system in a single crystal under
uniaxial tension.
\end{exercise}

\begin{exercise}
Derive the Taylor relation $\tau = \alpha G b \sqrt{\rho}$ from the
spacing between dislocations as obstacles to each other.
\end{exercise}

\begin{exercise}
Show that the von Mises criterion for general plastic deformation of
a polycrystalline metal requires at least five independent slip
systems, and identify why HCP metals deformed by basal slip alone do
not satisfy the criterion.
\end{exercise}

\begin{exercise}
Derive the Orowan stress $\tau_{\text{Orowan}} = G b/L$ for
dislocation bypass of incoherent precipitates of average spacing $L$.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the theoretical shear strength of FCC nickel ($G = 76\,
\text{GPa}$) and compare to the measured single-crystal critical
resolved shear stress ($\sim 5\,\text{MPa}$).
\end{exercise}

\begin{exercise}
Estimate the volume change of a $0.4\,\%$ C steel sample on quenching
from austenitising temperature, given the FCC austenite molar volume
and the BCT martensite molar volume. State the engineering consequence
for residual stress.
\end{exercise}

\begin{exercise}
Estimate the time required for hydrogen to diffuse out of a $10\,
\text{mm}$ thick steel section at $200\,\degC$ during a post-plating
bake, given $D_H \approx 10^{-9}\,\text{m}^2/\si{\second}$ at this
temperature. Compare to a $24\,\text{h}$ bake.
\end{exercise}

\begin{exercise}
Estimate the dislocation density in an annealed copper wire and in
a wire cold-drawn to $90\,\%$ area reduction, given that the
annealed yield strength is $50\,\text{MPa}$ and the drawn yield
strength is $400\,\text{MPa}$.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a two-dimensional discrete-dislocation simulation: a
distribution of edge dislocations in a square box, with an applied
shear stress. Integrate the equation of motion for each dislocation
under the combined applied stress and interaction stresses from all
other dislocations. Plot the stress-strain curve and extract the
Taylor relation coefficient $\alpha$.
\end{exercise}

\begin{exercise}
Implement a Monte Carlo grain-growth simulation in a Potts model.
Plot grain size versus annealing time and identify the parabolic
growth-law exponent.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Specify the alloy, heat treatment, and section thickness for a
bicycle frame that must support $200\,\text{N}$ of rider weight plus
$2\text{g}$ acceleration loading at the bottom bracket, with target
mass below $1.5\,\text{kg}$ for the frame. Compare an aluminium, a
steel, and a titanium option.
\end{exercise}

\begin{exercise}
Specify the alloy for a marine propeller shaft (rotating, sea-water
exposure, $5\,\text{m}$ length, $200\,\text{mm}$ diameter, $10\,\text{MW}$
power transmission). Identify the corrosion and fatigue requirements
and propose a candidate alloy and protection scheme.
\end{exercise}

\begin{exercise}
Specify the welding process and filler metal for joining a $25\,
\text{mm}$ thick stainless-steel pressure-vessel section in field
service. Identify the inspection requirements per ASME Section IX.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A heat-treated steel shaft has measured hardness $30\,\text{HRC}$,
substantially below the specification value of $45\,\text{HRC}$.
Identify the most likely metallurgical cause from the heat-treatment
process and propose two checks that would localise the cause.
\end{exercise}

\begin{exercise}
An aluminium casting shows significantly lower fatigue strength
than the alloy's specification value. Metallography shows coarse
silicon needles and porosity at the casting surface. Identify two
process modifications that would address each defect.
\end{exercise}

\begin{exercise}
A field-welded carbon-steel pipe joint shows a longitudinal crack at
the weld toe after one year of service. The pipe was operating in
buried, cathodically protected service. Identify the failure-mode
candidates and the inspection technique that distinguishes between
them.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
A high-strength bolt ($\sigma_{\text{UTS}} = 1200\,\text{MPa}$) fails
two days after installation in marine service. The fracture is
intergranular with no plastic deformation. Identify the failure mode
and propose the standard manufacturing-process change to eliminate
it.
\end{exercise}

\begin{exercise}
A welded austenitic stainless-steel chemical-process pipe (304L)
develops branched cracks at the inner surface in service at $80\,
\degC$ with chloride-containing process fluid. Identify the failure
mode and propose two material-substitution options.
\end{exercise}

\begin{exercise}
A rotating gas-turbine disk made of Inconel 718 develops a fatigue
crack at a fastener bore after $20\,000$ cycles. The bore was
cold-expanded during installation, leaving compressive residual
stress. Identify why the cold expansion is now insufficient to
prevent crack initiation and propose two changes to the disk's
inspection schedule.
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A colleague proposes substituting Ti-6Al-4V for AISI 4340 in a heavy
forging where the volume is large. Identify the cost, machinability,
and weldability differences that the substitution would introduce
and the application-specific factors that would make the substitution
beneficial or detrimental.
\end{exercise}

\begin{exercise}
A vendor claims that a new nanocrystalline aluminium alloy has yield
strength of $1500\,\text{MPa}$, exceeding 7075-T6 by a factor of
three, and that the alloy is suitable for aerospace structural use.
Identify the qualifications the vendor would need to document
(grain stability under service temperature, fatigue behaviour,
fracture toughness, weldability) and the inspection-level changes the
adoption would require.
\end{exercise}

\begin{exercise}
A student argues that all welded joints are weaker than the parent
metal. Identify the qualification this claim requires (it is
specific to the joint design and the welding process) and identify
a case in which the joint is at least as strong as the parent metal.
\end{exercise}
