\chapter{Viscous flow, boundary layers}
\label{vol03:ch12}

\epigraph{In many cases it is sufficient to find the path of the
fluid by assuming it inviscid; for here the influence of friction
is appreciable only in a very thin layer at the wall of the body.}{
Ludwig Prandtl, ``\"Uber Fl\"ussigkeitsbewegung bei sehr kleiner
Reibung'' (On the motion of fluid with very small friction),
Verhandlungen des III.\ Internationalen Mathematiker-Kongresses,
Heidelberg 1904 \cite[pp.~484--485]{paper:prandtl1904}.}

\chapmeta{Half-life: durable. Archetypes: transport, scaling, failure. Exercise target: 38. Project track: physical build (pressure-drop measurement on a household pipe). Hazard class: low. Reader path default: standard, with sections explicitly tagged. Named cases: Air France 447 high-altitude stall (2009).}

\noindent
The previous chapter treated the flow as inviscid: zero viscous
stress, no boundary layer, no separation, no turbulent dissipation.
Real fluids are viscous, and viscosity acts in a thin region
adjacent to solid surfaces and inside shear layers. This chapter
develops the Navier-Stokes equation for incompressible viscous
flow, derives the laminar exact solutions that close pipe flow
and Couette flow, develops Prandtl's boundary-layer theory, and
treats the transition to turbulence and its consequences for drag.
The reader leaves able to compute the Reynolds number for any
flow, identify whether the flow is laminar or turbulent, estimate
the boundary-layer thickness and skin-friction coefficient on a
surface, and recognise separation, stall, and the conditions under
which a viscous flow's drag departs from the inviscid prediction.

\section{Newton's law of viscosity}\pathtag{core}

A fluid resists shear deformation. If two parallel plates separated
by a thin layer of fluid are moved relative to one another at
constant velocity $U$ across a gap $h$, the steady-state velocity
profile in the gap is linear, and the shear stress required to
maintain the motion is proportional to the velocity gradient:
\[
\tau = \mu \frac{du}{dy}.
\]
The relation defines the \engterm{dynamic viscosity} $\mu$, a
material property of the fluid. A fluid obeying this linear
relation is a \engterm{Newtonian fluid}; air, water, and most
common liquids and gases at engineering conditions are Newtonian to
a very good approximation.

The dynamic viscosity has SI units of $\si{\pascal\second}$, or
equivalently $\si{\kilogram\per\meter\per\second}$. Representative
values at $20\,\degC$, current as of 2025 per
\textcite{text:white-fluid-mechanics}:

\begin{itemize}[leftmargin=*]
  \item Air: $\mu \approx 1.8\times 10^{-5}\,\si{\pascal\second}$.
  \item Water: $\mu \approx 1.0\times 10^{-3}\,\si{\pascal\second}$.
  \item Glycerin: $\mu \approx 1.5\,\si{\pascal\second}$.
  \item Honey (room temperature): $\mu \approx 10\,\si{\pascal\second}$,
    variable with sugar content and water content.
  \item Glass at room temperature: $\mu \sim 10^{21}\,\si{\pascal
    \second}$ (a solid, by the engineering definition of flow over
    relevant time scales).
\end{itemize}

Viscosity is strongly temperature dependent. Gases become more
viscous as they heat (faster molecular motion produces more
momentum transfer between layers); liquids become less viscous as
they heat (the molecular cohesion weakens). Engine oil drops by a
factor of $100$ in viscosity between $0\,\degC$ and $100\,\degC$,
which is why oil ratings carry temperature qualifiers (SAE 5W-30
specifies viscosity at both cold-start and operating temperatures).

The \engterm{kinematic viscosity} is
\[
\nu = \frac{\mu}{\rho},
\]
with SI units of $\si{\meter}^2/\si{\second}$. The kinematic
viscosity is the diffusivity of momentum in the fluid: it has the
same dimensions as a thermal diffusivity or a mass diffusivity, and
it characterises how rapidly velocity disturbances spread by
viscous action. For air at room conditions, $\nu \approx 1.5\times
10^{-5}\,\si{\meter}^2/\si{\second}$; for water, $\nu \approx
1.0\times 10^{-6}\,\si{\meter}^2/\si{\second}$. Water has lower
kinematic viscosity than air, despite higher dynamic viscosity,
because water is much denser.

\engterm{Non-Newtonian fluids} are those for which the
stress-strain-rate relation is nonlinear. Important classes
include:

\begin{itemize}[leftmargin=*]
  \item \engterm{Shear-thinning} (pseudoplastic): viscosity decreases
    with shear rate. Paint, blood, and many polymer solutions.
  \item \engterm{Shear-thickening} (dilatant): viscosity increases
    with shear rate. Concentrated cornstarch in water; certain
    body-armour fluids.
  \item \engterm{Bingham plastic}: behaves as a rigid solid below a
    yield stress, as a Newtonian fluid above. Toothpaste, drilling
    mud, fresh concrete.
  \item \engterm{Viscoelastic}: combines viscous and elastic
    responses. Polymer melts, biological tissues, silly putty.
\end{itemize}

Non-Newtonian fluids are the subject of \engterm{rheology}; we
treat them only in passing here. The Navier-Stokes equation of the
next section assumes a Newtonian fluid; rheologically complex
fluids require constitutive extensions.

\section{The Navier-Stokes equation}\pathtag{core}

Adding the viscous stress to the Euler equation of chapter 11
gives the \engterm{Navier-Stokes equation} for incompressible
Newtonian flow:
\[
\rho\frac{D\vec{u}}{Dt} = -\nabla p + \mu\nabla^2\vec{u} +
\rho\vec{g}.
\]
The viscous term $\mu\nabla^2\vec{u}$ replaces the surface-stress
contribution that the inviscid analysis ignored. The equation was
derived independently by Claude-Louis Navier in 1822 (from a
molecular argument) and by George Gabriel Stokes in 1845 (from a
continuum argument); both names are attached.

Combined with the incompressible continuity equation $\nabla\cdot
\vec{u} = 0$, the Navier-Stokes system is four equations in four
unknowns ($u_x$, $u_y$, $u_z$, $p$) in three space dimensions plus
time. The system is a coupled nonlinear partial differential
equation system whose general well-posedness for arbitrary initial
data is one of the seven Clay Millennium Prize problems
\cite{text:white-fluid-mechanics}; the engineering problem is not
mathematical existence but tractable analytical and numerical
solutions in specific geometries.

\engterm{From the stress tensor to the viscous term.} The viscous
term $\mu\nabla^2\vec u$ does not appear by assertion; it follows from
a constitutive law for the stress in a Newtonian fluid and the momentum
balance on a fluid element. We carry the derivation once, because every
later simplification in the chapter is a term dropped from the equation
this argument builds.

The state of stress at a point in a fluid is the symmetric tensor
$\sigma_{ij}$, whose component $\sigma_{ij}$ is the $j$-direction
traction on the face whose outward normal points along $i$
(figure~\ref{fig:vol03ch12:stress-cube}). Conservation of linear
momentum on a fluid element of density $\rho$ reads, in index notation
with the summation convention,
\[
\rho\frac{Du_i}{Dt} = \frac{\partial\sigma_{ij}}{\partial x_j} +
\rho g_i,
\]
the statement that the acceleration of the element equals the net
surface force per unit volume (the divergence of the stress) plus the
body force. This is Cauchy's equation of motion, and it holds for any
continuous medium, fluid or solid; the physics that distinguishes a
Newtonian fluid enters through the constitutive law that fixes
$\sigma_{ij}$.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-stress-cube}

Split the stress into an isotropic pressure part and a deviatoric
viscous part,
\[
\sigma_{ij} = -p\,\delta_{ij} + \tau_{ij},
\]
where $\delta_{ij}$ is the Kronecker delta and $\tau_{ij}$ is the
\engterm{viscous stress tensor} that vanishes when the fluid is at
rest. A fluid carries no shear when it is not deforming, so $\tau_{ij}$
must depend on the rate of deformation. The velocity-gradient tensor
$\partial u_i/\partial x_j$ splits into a symmetric strain-rate tensor
and an antisymmetric rotation-rate tensor,
\[
\frac{\partial u_i}{\partial x_j} = \underbrace{\frac12\!\left(
\frac{\partial u_i}{\partial x_j} + \frac{\partial u_j}{\partial x_i}
\right)}_{\displaystyle e_{ij}} + \underbrace{\frac12\!\left(
\frac{\partial u_i}{\partial x_j} - \frac{\partial u_j}{\partial x_i}
\right)}_{\displaystyle \Omega_{ij}},
\]
and a rigid rotation $\Omega_{ij}$ produces no internal stress, so
$\tau_{ij}$ depends on the strain-rate tensor $e_{ij}$ alone. The
Newtonian assumption is that this dependence is linear and isotropic:
the most general isotropic linear tensor function of $e_{ij}$ is
\[
\tau_{ij} = 2\mu\,e_{ij} + \lambda\,e_{kk}\,\delta_{ij},
\]
with two material constants, the dynamic viscosity $\mu$ and the second
viscosity $\lambda$. The trace $e_{kk} = \partial u_k/\partial x_k =
\nabla\cdot\vec u$ is the volumetric dilatation rate, which vanishes for
an incompressible fluid. Setting $\nabla\cdot\vec u = 0$ drops the
$\lambda$ term, leaving
\[
\tau_{ij} = 2\mu\,e_{ij} = \mu\!\left(\frac{\partial u_i}{\partial x_j}
+ \frac{\partial u_j}{\partial x_i}\right).
\]
This is the tensor form of Newton's law of viscosity; the simple shear
$\tau = \mu\,du/dy$ of the opening section is its single off-diagonal
component for a unidirectional flow.

Substitute the constitutive law back into Cauchy's equation. The
divergence of the stress is
\[
\frac{\partial\sigma_{ij}}{\partial x_j} = -\frac{\partial p}{\partial
x_i} + \mu\frac{\partial}{\partial x_j}\!\left(\frac{\partial u_i}{
\partial x_j} + \frac{\partial u_j}{\partial x_i}\right) =
-\frac{\partial p}{\partial x_i} + \mu\frac{\partial^2 u_i}{\partial
x_j\partial x_j} + \mu\frac{\partial}{\partial x_i}\!\left(
\frac{\partial u_j}{\partial x_j}\right),
\]
where the constant $\mu$ has been taken outside the derivative. The
last term carries $\partial u_j/\partial x_j = \nabla\cdot\vec u = 0$
and so vanishes for incompressible flow, leaving $\mu\,\partial^2
u_i/\partial x_j\partial x_j = \mu\nabla^2 u_i$. The momentum balance
becomes
\[
\rho\frac{Du_i}{Dt} = -\frac{\partial p}{\partial x_i} + \mu\nabla^2
u_i + \rho g_i,
\]
which is the incompressible Navier-Stokes equation in components, the
vector form stated above. The viscous term is the divergence of the
viscous stress, and its Laplacian shape is a consequence of the linear
isotropic constitutive law plus incompressibility, not an independent
postulate. The assumptions used, in order, are the continuum momentum
balance, the decomposition of stress into pressure and deviatoric
parts, the dependence of viscous stress on strain rate alone, linearity
and isotropy of that dependence, and incompressibility. Each later
section of the chapter relaxes or specialises one of these: the
exact-solution sections drop the convective acceleration on the left,
the boundary-layer section drops one viscous term by a scaling
argument, and the turbulence section averages the whole equation and
confronts the new stress the averaging leaves behind.

\begin{mastery}
\textbf{The vorticity form and why viscosity diffuses spin.} Taking the
curl of the incompressible Navier-Stokes equation eliminates the
pressure (the curl of a gradient is zero) and produces the
\engterm{vorticity transport equation} for $\vec\omega = \nabla\times
\vec u$,
\[
\frac{D\vec\omega}{Dt} = (\vec\omega\cdot\nabla)\vec u + \nu\nabla^2
\vec\omega.
\]
The pressure has vanished, so the equation reads as a balance of two
mechanisms acting on vorticity. The term $(\vec\omega\cdot\nabla)
\vec u$ is \engterm{vortex stretching}: a vortex tube stretched by the
strain field intensifies its spin, conserving angular momentum like a
spinning skater pulling in her arms, and this term exists only in three
dimensions, which is the structural reason three-dimensional turbulence
cascades energy to small scales while two-dimensional turbulence does
not. The term $\nu\nabla^2\vec\omega$ is \engterm{viscous diffusion of
vorticity}: viscosity spreads spin from regions of high vorticity to
low, exactly as conduction spreads heat, with the kinematic viscosity
$\nu$ playing the role of a diffusivity for spin. In a flow started from
rest the only source of vorticity is the no-slip condition at solid
walls, which injects vorticity at the surface; viscosity then diffuses
it outward into the boundary layer, and the convection carries it
downstream. This is the deepest statement of what a boundary layer is:
the region into which wall-generated vorticity has diffused, of
thickness $\delta \sim \sqrt{\nu t}$ in time $t$ or
$\delta \sim \sqrt{\nu x/U_\infty}$ at distance $x$, the same
$\sqrt{\nu t}$ law as Stokes's first problem and as heat diffusion. The
inviscid flow of chapter~11 carries no mechanism to generate or destroy
vorticity, which is why an inviscid analysis of flow started from rest
stays irrotational and predicts zero drag; viscosity, however small, is
the term that lets vorticity enter at the wall and so is the term that
produces drag. The vorticity equation thus packages the whole
chapter's physics in one line: stretching builds the small scales of
turbulence, diffusion sets the boundary-layer thickness, and the
wall is the source.
\end{mastery}

The Navier-Stokes equation is the foundation of nearly all of fluid
mechanics from chapter 12 onward. The strategy of analytical fluid
mechanics is to find configurations in which the equation
simplifies enough to admit a closed-form solution (the exact
solutions of section 12.4), find configurations in which one term
is small and can be neglected (the boundary-layer approximation of
section 12.5), or solve the equation numerically on a discretised
grid (the computational-fluid-dynamics methods of section 12.7).

\engterm{Non-dimensionalisation}. The Navier-Stokes equation has a
characteristic non-dimensional form. Introduce a characteristic
length scale $L$, a characteristic velocity $U$, a characteristic
time scale $L/U$, and a characteristic pressure $\rho U^2$. Let
$\vec{u}^* = \vec{u}/U$, $\vec{x}^* = \vec{x}/L$, $t^* = tU/L$,
$p^* = (p - p_\infty)/(\rho U^2)$. The non-dimensional
Navier-Stokes equation becomes
\[
\frac{D\vec{u}^*}{Dt^*} = -\nabla^* p^* + \frac{1}{\text{Re}}
\nabla^{*2}\vec{u}^* + \frac{1}{\text{Fr}^2}\hat{g},
\]
where
\[
\text{Re} = \frac{\rho U L}{\mu} = \frac{U L}{\nu}, \quad
\text{Fr} = \frac{U}{\sqrt{gL}}.
\]
The \engterm{Reynolds number} Re is the ratio of inertial to
viscous forces; the \engterm{Froude number} Fr is the ratio of
inertial to gravitational forces. Two flows with the same
non-dimensional geometry and the same Re (and Fr, if gravity is
relevant) have the same non-dimensional velocity and pressure
fields. This is the principle of \engterm{dynamic similarity}: it
is why small wind-tunnel models tested at appropriate Reynolds
number predict the behaviour of full-scale aircraft, ship hulls,
and bridges.

\begin{principle}[The Reynolds number sets the flow regime]
The Reynolds number $\text{Re} = \rho U L/\mu = U L/\nu$ measures
the ratio of inertial to viscous forces in a flow. Low Re ($\ll 1$)
corresponds to creeping flow dominated by viscosity; intermediate
Re ($\sim 1$ to $\sim 10^3$) to laminar viscous flow; high Re
($\gtrsim 10^4$ to $10^5$, geometry dependent) to turbulent flow.
The Reynolds number is the single most important dimensionless
parameter in viscous fluid mechanics, and the working engineer
computes it before doing any other analysis.
\end{principle}

Reynolds's original 1883 experiment on the transition from laminar
to turbulent flow in a glass-walled pipe used a dye filament
injected on the pipe centreline. At low flow rates the filament
remained straight and parallel to the pipe axis, indicating
laminar flow; at high flow rates the filament dispersed across
the pipe, indicating turbulent mixing. Reynolds identified the
combination $\rho U D/\mu$ (with $D$ the pipe diameter) as the
parameter governing the transition, and reported a critical value
near $2000$ for the onset of intermittent turbulence and near
$13{,}000$ for fully developed turbulence under his conditions
\cite[pp.~935--982]{paper:reynolds1883}. The result is the
empirical basis of the laminar-turbulent transition criterion in
pipe flow.

\section{Pipe flow: laminar and turbulent}\pathtag{core}

A steady flow of incompressible Newtonian fluid through a long
circular pipe of diameter $D$ at low Reynolds number admits an
exact solution. With the axial coordinate $z$ along the pipe and
the radial coordinate $r$ from the centreline, the steady
Navier-Stokes equation in cylindrical coordinates gives, under
the assumptions of fully developed flow ($\partial u_z/\partial z
= 0$, $u_r = 0$, $u_\theta = 0$) and constant axial pressure
gradient $dp/dz = -G$,
\[
\mu\frac{1}{r}\frac{d}{dr}\left(r\frac{du_z}{dr}\right) = -G.
\]
Integrating twice with boundary conditions $u_z(r=R) = 0$ (no-slip
at the wall, with $R = D/2$) and $du_z/dr |_{r=0} = 0$ (symmetry
at the centreline),
\[
u_z(r) = \frac{G(R^2 - r^2)}{4\mu}.
\]
This is the \engterm{Hagen-Poiseuille velocity profile}, a
parabola with the maximum on the axis and zero at the wall
(figure~\ref{fig:vol03ch12:poiseuille-profile}). The
centreline velocity is $u_{\max} = GR^2/(4\mu)$, and the
cross-section-averaged velocity is half the maximum:
\[
\bar{u} = \frac{GR^2}{8\mu}.
\]
\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-poiseuille-profile}

The volume flow rate is $Q = \pi R^2 \bar{u} = \pi R^4 G/(8\mu)$,
giving the \engterm{Hagen-Poiseuille law}
\[
Q = \frac{\pi D^4 (-\Delta p)}{128 \mu L},
\]
where $L$ is the pipe length and $\Delta p = p_2 - p_1 = -GL$ is
the (negative) pressure change along the length. The flow rate
scales as $D^4$: doubling the diameter increases the flow rate by
sixteen for the same pressure drop. The same scaling explains why
arterial flow drops sharply when arteries narrow; a $50\,\%$
reduction in lumen diameter reduces the flow rate by a factor of
$16$.

The \engterm{Darcy-Weisbach equation} casts the pressure drop in
terms of a \engterm{friction factor} $f$:
\[
\Delta p = f\frac{L}{D}\,\frac{\rho \bar{u}^2}{2}.
\]
For laminar pipe flow, the friction factor is
\[
f = \frac{64}{\text{Re}_D}, \quad \text{Re}_D = \frac{\rho \bar{u} D}{\mu}.
\]
The relation is exact under the Hagen-Poiseuille assumptions and
holds for $\text{Re}_D \lesssim 2300$.

Above $\text{Re}_D \approx 2300$, the laminar pipe flow becomes
unstable to perturbations and transitions to turbulence. The
turbulent regime begins with intermittent slugs of disturbed flow
at $\text{Re}_D \sim 2300$ to $4000$ and reaches fully developed
turbulence above $\text{Re}_D \sim 4000$. In the turbulent regime
the velocity profile is much flatter than the laminar parabola
(with the velocity dropping rapidly only in a thin near-wall
region) and the friction factor depends on the relative pipe-wall
roughness $\epsilon/D$ in addition to $\text{Re}_D$. The contrast
between the two profile shapes is drawn in
figure~\ref{fig:vol03ch12:laminar-turbulent}: the turbulent
profile fills the core almost uniformly and falls steeply through
a thin near-wall region, so it carries a far larger wall velocity
gradient and a larger friction factor than the laminar parabola at
the same flow rate.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-laminar-turbulent-profiles}

The \engterm{Moody chart}
\cite{paper:moody-1944,text:white-fluid-mechanics}, shown
schematically in figure~\ref{fig:vol03ch12:moody},
plots $f$ versus $\text{Re}_D$ for various values of $\epsilon/D$,
condensing decades of empirical and analytical work into a single
working tool. For fully turbulent flow, the
implicit Colebrook-White equation \cite{paper:colebrook-1939}
\[
\frac{1}{\sqrt{f}} = -2\log_{10}\left(\frac{\epsilon/D}{3.7} +
\frac{2.51}{\text{Re}_D\sqrt{f}}\right)
\]
gives the friction factor; explicit approximations (Swamee-Jain,
Haaland) are used in practice for direct calculation. The Haaland
form
\[
\frac{1}{\sqrt{f}} = -1.8\log_{10}\!\left[
\left(\frac{\epsilon/D}{3.7}\right)^{1.11} +
\frac{6.9}{\text{Re}_D}\right]
\]
is explicit in $f$ and stays within a few percent of the
Colebrook-White root across the engineering range, which is why it
is the workhorse for direct calculation; the solver in
\texttt{code/moody.py} carries both forms and the friction-factor
table behind figure~\ref{fig:vol03ch12:moody} lives in
\texttt{data/moody\_friction.csv}. The chart
shows three regimes:

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-moody-chart}

\begin{itemize}[leftmargin=*]
  \item \engterm{Laminar regime} ($\text{Re}_D < 2300$): $f$
    depends only on $\text{Re}_D$, with $f = 64/\text{Re}_D$.
  \item \engterm{Transitional regime} ($2300 < \text{Re}_D <
    4000$): unstable, intermittent, no single-curve description.
  \item \engterm{Turbulent regime} ($\text{Re}_D > 4000$): $f$
    depends on $\text{Re}_D$ and $\epsilon/D$. At high enough
    Reynolds number (the \engterm{fully rough regime}), $f$
    depends only on $\epsilon/D$ and not on $\text{Re}_D$.
\end{itemize}

The Moody chart is the working tool for pipe-flow design in every
hydraulics, HVAC, and process-engineering application.

\section{Couette flow and other exact laminar solutions}\pathtag{standard}

\engterm{Couette flow}. A Newtonian fluid between two infinite
parallel plates, one at $y = 0$ and stationary, the other at
$y = h$ moving at velocity $U$ in the $x$-direction. In steady
state with no pressure gradient and gravity ignored, the
Navier-Stokes equation reduces to $d^2 u_x/dy^2 = 0$, with
solution
\[
u_x(y) = \frac{U y}{h}.
\]
The velocity profile is linear; the shear stress is constant
across the gap, $\tau = \mu U/h$. The force per unit area required
to move the upper plate is the same shear stress, which appears as
the viscous drag on the plate. This is the canonical configuration
for a rotating viscometer (concentric cylinders with a small gap),
which measures $\mu$ directly from the torque required to
maintain a known rotation rate.

\engterm{Combined Couette-Poiseuille flow}. The same geometry with
a pressure gradient added gives a velocity profile that is the
superposition of the linear Couette profile and a parabolic
Poiseuille component. For favourable pressure gradient
($dp/dx < 0$ with flow in the $+x$-direction), the profile is a
forward-leaning shape with maximum velocity above the moving
plate's level; for adverse pressure gradient, the profile has a
reverse-flow region near the stationary plate when the pressure
gradient is large enough. The three profiles are drawn together in
figure~\ref{fig:vol03ch12:couette-poiseuille}. The
adverse-pressure-gradient case is
the simplest analytical model of \engterm{flow reversal} that
underlies boundary-layer separation in section 12.6.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-couette-poiseuille-profiles}

\engterm{Stokes's first problem}. An infinite plate impulsively
accelerated to constant velocity $U$ at $t = 0$ in a fluid at
rest. The momentum diffuses out from the plate into the fluid; the
velocity profile at time $t$ is
\[
u_x(y, t) = U\,\text{erfc}\left(\frac{y}{2\sqrt{\nu t}}\right),
\]
where erfc is the complementary error function. The velocity
disturbance penetrates to a depth $\delta(t) \sim \sqrt{\nu t}$;
the relation $\delta \sim \sqrt{\nu t}$ is the canonical viscous
diffusion length, the spatial analogue of the heat-diffusion
relation of chapter 4 of Volume IV.

\engterm{Stokes flow past a sphere}. A sphere of radius $a$
moving at velocity $U$ through a viscous fluid at very low
Reynolds number (Re $\ll 1$, the \engterm{creeping flow} regime)
experiences a drag
\[
F_D = 6\pi \mu a U,
\]
the \engterm{Stokes drag law} \cite{paper:ch02t3b-stokes-1851}. The law underlies
the design of falling-ball viscometers, the settling velocity of
small particles in suspensions, and the propulsion of microorganisms.
The Stokes drag holds for Re $< 1$ on the sphere; for larger Re,
inertial corrections (Oseen, 1910) and eventual wake separation
modify the relation. The drag coefficient $C_D = F_D/(\tfrac{1}{2}
\rho U^2 \pi a^2) = 24/\text{Re}$ in the Stokes regime, $\sim 0.4$
at $\text{Re} = 10^5$ (the canonical drag-crisis transition), and
varies non-monotonically across the transition (the topic of
section 12.6). The full sphere-drag curve across nine decades of
Reynolds number is drawn in
figure~\ref{fig:vol03ch12:sphere-drag}; the tabulated standard
curve is in \texttt{data/sphere\_drag\_curve.csv} and the
iterative settling-velocity solver that uses it is in
\texttt{code/sphere\_terminal\_velocity.py}.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-sphere-drag-curve}

A settling or terminal-velocity calculation needs only the part of
this curve that brackets the answer, and the working device is to
know which of three regimes the body sits in. The Stokes line
$C_D = 24/\text{Re}$ holds below $\text{Re}\approx1$; the
Schiller-Naumann correlation $C_D = (24/\text{Re})(1 + 0.15\,
\text{Re}^{0.687})$ bridges the intermediate range; and above
$\text{Re}\approx10^3$ the coefficient settles onto the Newton
plateau $C_D\approx0.44$ until the drag crisis intervenes. The three
regimes and their boundaries are drawn in
figure~\ref{fig:vol03ch12:terminal-velocity-regimes}; the
regime check that worked examples~3 and~8 carry is a reading of which
segment of this curve applies.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-terminal-velocity-regimes}

\section{Dimensional analysis and the Pi theorem}\pathtag{standard}

The non-dimensional Navier-Stokes equation produced the Reynolds and
Froude numbers from the equation itself. The same groups, and the form
of the friction factor and the drag coefficient, follow from
dimensional analysis alone, without writing the governing equation.
The method is the Buckingham Pi theorem, and it earns its place here
because the relations the chapter uses, the friction factor
$f(\text{Re},\epsilon/D)$ and the drag coefficient $C_D(\text{Re})$,
are exactly the dimensionless groupings the theorem predicts.

\engterm{The Pi theorem.} A physical relation among $n$ dimensional
quantities that involves $k$ independent base dimensions can be
rewritten as a relation among $n - k$ independent dimensionless groups.
The base dimensions of mechanics are mass $M$, length $L$, and time
$T$, so $k = 3$ for the flows of this chapter. The working steps are to
list the variables, count the base dimensions they span, choose $k$
repeating variables that together carry all the base dimensions, and
form the remaining $n - k$ variables into dimensionless groups by
dividing each by a product of powers of the repeating variables.

\engterm{Pipe-flow pressure drop.} Take the pressure drop per unit
length in a pipe to depend on the variables that physically set it: the
mean velocity $\bar u$, the diameter $D$, the density $\rho$, the
viscosity $\mu$, and the wall roughness $\epsilon$. The relation is
\[
\frac{\Delta p}{L} = \phi(\bar u, D, \rho, \mu, \epsilon).
\]
Counting, $n = 6$ quantities ($\Delta p/L$, $\bar u$, $D$, $\rho$,
$\mu$, $\epsilon$) span $k = 3$ base dimensions, so the theorem
predicts $n - k = 3$ dimensionless groups. Choose $\bar u$, $D$, $\rho$
as the repeating variables; they carry $L T^{-1}$, $L$, and $M L^{-3}$
respectively, which together span $M$, $L$, $T$. Form the groups by
making each remaining variable dimensionless. The pressure-drop group
is
\[
\Pi_1 = \frac{(\Delta p/L)\,D}{\rho\,\bar u^2},
\]
which is the friction factor up to a constant (the Darcy-Weisbach
definition packages it as $f = (\Delta p/L)\,D/(\tfrac12\rho\bar u^2)$,
the same group with the factor of two folded in). The viscosity group
is
\[
\Pi_2 = \frac{\mu}{\rho\,\bar u\,D} = \frac{1}{\text{Re}_D},
\]
the inverse Reynolds number, and the roughness group is the geometric
ratio
\[
\Pi_3 = \frac{\epsilon}{D}.
\]
The six-variable dimensional relation therefore collapses to a relation
among three groups,
\[
f = F\!\left(\text{Re}_D,\ \frac{\epsilon}{D}\right),
\]
which is exactly the statement the Moody chart plots. Dimensional
analysis does not give the function $F$; that comes from theory
(Hagen-Poiseuille fixes $f = 64/\text{Re}_D$ in the laminar branch) or
from experiment (Colebrook-White fits the turbulent branch). What it
gives is the shape of the answer: the friction factor depends on the
flow only through the Reynolds number and the relative roughness, so a
single chart in two variables covers every pipe, fluid, and flow rate.

\engterm{The Euler number and the drag coefficient.} The same
machinery applied to the force on a body returns the drag coefficient.
A drag force $F_D$ depending on $\rho$, $U$, a body size $L$, and $\mu$
has $n = 5$ quantities in $k = 3$ dimensions, so $n - k = 2$ groups.
With $\rho$, $U$, $L$ repeating, the force group is
\[
\Pi_1 = \frac{F_D}{\rho\,U^2\,L^2} = \frac{F_D}{\rho\,U^2 A}\cdot
\text{(geometric constant)},
\]
the drag coefficient $C_D$ up to the conventional factor of $\tfrac12$
and the choice of reference area $A\propto L^2$, and the second group
is again $\mu/(\rho U L) = 1/\text{Re}$. So $C_D = G(\text{Re})$ for a
fixed body shape, the statement section~12.5 makes when it tabulates
$C_D$ against Reynolds number. The pressure group $\Delta p/(\rho U^2)$
that recurs in both analyses is the \engterm{Euler number}
$\text{Eu} = \Delta p/(\rho U^2)$, the ratio of pressure forces to
inertial forces; the friction factor and the drag coefficient are both
Euler numbers built on different reference pressures. The reason the
factor $\tfrac12\rho U^2$ appears in every coefficient definition in
the chapter is that it is the dynamic pressure, the inertial pressure
scale the Pi theorem selects.

\engterm{Why dynamic similarity works.} The Pi theorem is the formal
reason a scale model predicts a prototype. Two geometrically similar
flows with the same values of every dimensionless group have the same
value of every dimensionless output, because the output is a function
of those groups and nothing else. Matching the Reynolds number (and,
where they matter, the Froude, Mach, or Weber number) between model and
prototype makes the dimensionless force coefficients equal, which is
the principle worked example~6 applied to a submarine model and worked
example~17 below applies to a wind-tunnel test. The limit of the method
is that a model can rarely match every group at once: a ship model
matching Froude number runs at the wrong Reynolds number, which is why
ship drag is split into a friction part and a wave-making part and each
is scaled by its own group, the procedure worked example~12 named for
Froude.

\section{Boundary-layer theory}\pathtag{core}

Ludwig Prandtl's 1904 Heidelberg paper
\cite{paper:prandtl1904} resolved a paradox that had stood since
d'Alembert: an inviscid analysis of flow past a body gives zero
drag, but real bodies in real flow experience large drag. Prandtl
observed that viscosity, however small, cannot be neglected in a
thin region adjacent to the surface where the no-slip condition
$\vec{u} = 0$ at the wall forces a steep velocity gradient. He
named this region the \engterm{boundary layer} and showed that
inside it the Navier-Stokes equations simplify to a system that is
analytically tractable.

For a flat plate at zero angle of attack in a uniform freestream
$U_\infty$, the boundary-layer assumption is that the layer
thickness $\delta(x)$ is much smaller than the streamwise distance
$x$. Inside the layer, the streamwise momentum equation becomes
\[
u\frac{\partial u}{\partial x} + v\frac{\partial u}{\partial y} =
-\frac{1}{\rho}\frac{\partial p}{\partial x} + \nu\frac{\partial^2
u}{\partial y^2},
\]
with the wall-normal momentum equation reducing to $\partial
p/\partial y = 0$ (the pressure does not vary across the layer).
The pressure inside the boundary layer is therefore set by the
inviscid outer flow.

\engterm{The boundary-layer equations from order-of-magnitude
scaling.} The reduction from the two-dimensional Navier-Stokes
equations to the single streamwise equation above is an exercise in
estimating the size of each term. Start from the steady,
two-dimensional, incompressible Navier-Stokes equations with no body
force,
\[
u\frac{\partial u}{\partial x} + v\frac{\partial u}{\partial y} =
-\frac1\rho\frac{\partial p}{\partial x} + \nu\!\left(
\frac{\partial^2 u}{\partial x^2} + \frac{\partial^2 u}{\partial y^2}
\right),
\]
\[
u\frac{\partial v}{\partial x} + v\frac{\partial v}{\partial y} =
-\frac1\rho\frac{\partial p}{\partial y} + \nu\!\left(
\frac{\partial^2 v}{\partial x^2} + \frac{\partial^2 v}{\partial y^2}
\right),
\]
together with continuity $\partial u/\partial x + \partial v/\partial y
= 0$. Assign each variable its characteristic scale: streamwise
distance varies over the plate length $L$, wall-normal distance over
the layer thickness $\delta$, and the streamwise velocity over the
freestream value $U_\infty$ (figure~\ref{fig:vol03ch12:bl-scaling}).
The boundary-layer assumption is the single inequality
$\delta \ll L$.

Continuity fixes the wall-normal velocity scale. The streamwise term
$\partial u/\partial x$ scales as $U_\infty/L$, and for continuity to
balance, the wall-normal term $\partial v/\partial y$ must be the same
size, so with $v$ varying over $\delta$,
\[
\frac{v}{\delta} \sim \frac{U_\infty}{L} \implies v \sim U_\infty
\frac{\delta}{L} \ll U_\infty.
\]
The cross-stream velocity is smaller than the streamwise velocity by
the aspect ratio $\delta/L$. Carry these scales through the streamwise
momentum equation term by term. The two convective terms are each of
order
\[
u\frac{\partial u}{\partial x} \sim \frac{U_\infty^2}{L}, \qquad
v\frac{\partial u}{\partial y} \sim U_\infty\frac{\delta}{L}\cdot
\frac{U_\infty}{\delta} = \frac{U_\infty^2}{L},
\]
so both survive at the same order. The two viscous terms are
\[
\nu\frac{\partial^2 u}{\partial x^2} \sim \frac{\nu U_\infty}{L^2},
\qquad
\nu\frac{\partial^2 u}{\partial y^2} \sim \frac{\nu U_\infty}{\delta^2},
\]
and their ratio is $(\delta/L)^2 \ll 1$: the streamwise diffusion is
smaller than the wall-normal diffusion by the square of the aspect
ratio, so it is dropped. The wall-normal diffusion is retained only if
it is comparable to the convection that drives the layer, which sets
the thickness:
\[
\frac{U_\infty^2}{L} \sim \frac{\nu U_\infty}{\delta^2} \implies
\frac{\delta}{L} \sim \frac{1}{\sqrt{U_\infty L/\nu}} =
\frac{1}{\sqrt{\text{Re}_L}}.
\]
The boundary layer is thin precisely because the Reynolds number is
large, and the same balance is what worked example~3 of the derivation
exercises reproduces from the bare scaling. Applying the identical
estimates to the wall-normal momentum equation shows every term in it
is smaller than the streamwise terms by a factor $\delta/L$ or its
square, except the pressure derivative, which leaves
\[
\frac{\partial p}{\partial y} \approx 0
\]
as the only surviving balance: the pressure is constant across the
layer to leading order, so $p(x)$ is impressed on the layer by the
outer inviscid flow through the edge velocity $U_e(x)$ via the
Bernoulli relation $dp/dx = -\rho U_e\,dU_e/dx$. The streamwise
equation, with the streamwise diffusion gone, is the boundary-layer
momentum equation stated above. The reduction has turned an elliptic
system, in which information travels in every direction, into a
parabolic one that marches downstream from an inlet profile, which is
why the boundary-layer equations are solved by marching while the full
Navier-Stokes equations are not.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-bl-scaling}

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-boundary-layer-growth}

For a flat plate at zero pressure gradient ($\partial p/\partial x
= 0$), Heinrich Blasius (1908) solved the boundary-layer equations
by a similarity transformation
\cite{text:schlichting-boundary-layer}. The boundary-layer
thickness grows as
\[
\delta(x) \approx 5.0\,\sqrt{\frac{\nu x}{U_\infty}} =
\frac{5.0\,x}{\sqrt{\text{Re}_x}},
\]
where $\text{Re}_x = U_\infty x/\nu$ is the local Reynolds number
based on streamwise distance. The local skin-friction coefficient
is
\[
C_f(x) = \frac{\tau_w(x)}{\tfrac{1}{2}\rho U_\infty^2} =
\frac{0.664}{\sqrt{\text{Re}_x}},
\]
where $\tau_w$ is the wall shear stress. The total drag on a plate
of length $L$ is obtained by integrating $\tau_w(x)$:
\[
F_D = \tfrac{1}{2}\rho U_\infty^2 \cdot L \cdot \bar{C}_f, \quad
\bar{C}_f = \frac{1.328}{\sqrt{\text{Re}_L}}.
\]
These are the Blasius solutions; they form the analytical
foundation of boundary-layer engineering. The shape of the
underlying similarity profile is shown in
figure~\ref{fig:vol03ch12:blasius}: plotting $u/U_\infty$ against
the similarity variable $\eta = y\sqrt{U_\infty/(\nu x)}$ collapses
the profile at every streamwise station onto a single curve, with
the velocity reaching $99\%$ of the freestream at
$\eta_{99}\approx5.0$. The tabulated profile is in
\texttt{data/blasius\_profile.csv}, and the shooting-method solver
that reproduces it (and the wall-shear constant $f''(0)\approx
0.33206$ that fixes $C_f$) is in \texttt{code/blasius.py}.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-blasius-profile}

\begin{mastery}
\textbf{The Blasius similarity solution.} For a flat plate at zero
pressure gradient the boundary-layer equations admit a similarity
reduction: the streamwise velocity depends on $x$ and $y$ only
through the combination $\eta = y\sqrt{U_\infty/(\nu x)}$. Introduce
a stream function $\psi = \sqrt{\nu x\,U_\infty}\,f(\eta)$, so that
$u = \partial\psi/\partial y = U_\infty f'(\eta)$ and $v =
-\partial\psi/\partial x$. Substituting into the streamwise
momentum equation $u\,u_x + v\,u_y = \nu\,u_{yy}$ collapses the
partial differential equation to the ordinary differential equation
\[
2f''' + f f'' = 0,
\]
with boundary conditions $f(0) = 0$ (no penetration), $f'(0) = 0$
(no-slip), and $f'(\infty) = 1$ (match the freestream). The equation
is third order and nonlinear with a split boundary condition, so no
elementary closed form exists; it is integrated numerically. The
standard route is a shooting method: guess the wall curvature
$f''(0)$, integrate to large $\eta$, and adjust the guess until
$f'(\infty)\to 1$. The converged value $f''(0) = 0.33206$ fixes the
wall shear stress $\tau_w = \mu (\partial u/\partial y)_{y=0} =
0.332\,\rho U_\infty^2/\sqrt{\text{Re}_x}$, and hence the
skin-friction coefficient $C_f = 0.664/\sqrt{\text{Re}_x}$. The
$\sqrt{\nu x/U_\infty}$ scaling of the thickness, the
$1/\sqrt{\text{Re}_x}$ scaling of $C_f$, and the $1/\sqrt{
\text{Re}_L}$ scaling of the integrated drag all descend from this
one similarity reduction.
\end{mastery}

\engterm{Displacement thickness}. The boundary layer reduces the
mass flux past a surface compared to what an inviscid flow would
deliver. The \engterm{displacement thickness} $\delta^*$ is the
distance the outer streamline would have to be moved outward to
recover the inviscid mass flux:
\[
\delta^* = \int_0^\infty \left(1 - \frac{u}{U_\infty}\right) dy.
\]
For the Blasius profile, $\delta^*/\delta \approx 0.345$, so the
displacement thickness is about a third of the full
boundary-layer thickness. The displacement thickness is the
correction to the inviscid streamline geometry needed to account
for the boundary layer in coupled analyses: a wind-tunnel test
section whose walls grow a boundary layer has an effective
cross-section that shrinks by $\delta^*$ on each wall, so the
freestream accelerates down the tunnel unless the walls are
diverged to compensate, and an airfoil analysis that wants the
real pressure distribution adds $\delta^*(x)$ to the surface and
re-solves the inviscid flow over the thickened shape. The three
length scales the profile carries, the full thickness $\delta$, the
displacement thickness $\delta^*$, and the momentum thickness
$\theta$ introduced next, are read off one velocity profile as the
total height, the velocity-deficit area, and the momentum-deficit
area respectively; figure~\ref{fig:vol03ch12:thicknesses} shows the
three on a single profile, and worked example~10 evaluates them for
a sinusoidal profile.

\engterm{Momentum thickness}. The \engterm{momentum thickness}
$\theta$ is
\[
\theta = \int_0^\infty \frac{u}{U_\infty}\left(1 - \frac{u}{U_
\infty}\right) dy.
\]
The Kármán momentum integral relates $\theta$ to the wall shear
stress:
\[
\frac{d\theta}{dx} = \frac{C_f(x)}{2}
\]
(for zero pressure gradient). This integral form is the
working tool for approximate boundary-layer calculations when the
exact Blasius solution does not apply.

\engterm{Deriving the momentum integral.} The relation is obtained by
integrating the boundary-layer momentum equation across the layer, and
carrying the pressure gradient through gives the general form that
worked examples~7 and~14 use. Write the boundary-layer momentum
equation with the edge pressure expressed through the edge velocity,
$-\tfrac1\rho\,dp/dx = U_e\,dU_e/dx$, and add to it the continuity
equation multiplied by $(u - U_e)$, a step that lets the convective
terms be written as a single derivative:
\[
\frac{\partial}{\partial x}\big[u(u - U_e)\big] +
\frac{\partial}{\partial y}\big[v(u - U_e)\big] +
(u - U_e)\frac{dU_e}{dx} = \nu\frac{\partial^2 u}{\partial y^2}.
\]
Integrate from the wall $y = 0$ to a height $y = h$ above the layer
where $u = U_e$ and the velocity gradient vanishes. The second term
integrates to zero at both limits ($u - U_e = 0$ at the edge and
$v = 0$ at the wall). The right side integrates to $-\nu\,(\partial
u/\partial y)_{y=0} = -\tau_w/\rho$, the wall shear with a sign set by
the inward integration. The first and third terms combine into
\[
\frac{d}{dx}\int_0^h u(u - U_e)\,dy + \frac{dU_e}{dx}\int_0^h
(u - U_e)\,dy = -\frac{\tau_w}{\rho}.
\]
Recognise the two integrals as the momentum and displacement
thicknesses: $\int_0^h u(U_e - u)\,dy = U_e^2\,\theta$ and $\int_0^h
(U_e - u)\,dy = U_e\,\delta^*$. Substituting and dividing by $U_e^2$
gives the \engterm{Kármán momentum-integral equation} in its general
form,
\[
\frac{d\theta}{dx} + (2\theta + \delta^*)\frac{1}{U_e}\frac{dU_e}{dx}
= \frac{C_f}{2},
\]
usually written with the shape factor $H = \delta^*/\theta$ as
\[
\frac{d\theta}{dx} + (2 + H)\frac{\theta}{U_e}\frac{dU_e}{dx} =
\frac{C_f}{2}.
\]
A favourable edge gradient ($dU_e/dx > 0$) adds a sink term that holds
the momentum thickness down; an adverse gradient ($dU_e/dx < 0$) feeds
the layer and grows $\theta$ faster, the integral statement of the
thickening that precedes separation. Setting $dU_e/dx = 0$ recovers the
flat-plate result $d\theta/dx = C_f/2$. The equation has one relation
in three unknowns ($\theta$, $H$, $C_f$), so it closes only with an
assumed velocity profile, which is what the cubic profile of worked
example~7, the sinusoidal profile of worked example~10, and the
Pohlhausen quartic of worked example~14 supply, and what Thwaites'
correlation supplies through a fitted universal function.

\begin{mastery}
\textbf{Thwaites' method: the integral layer with an arbitrary
pressure gradient.} The Kármán integral becomes a working predictor
for laminar layers under any pressure gradient through Thwaites'
correlation (1949). Carry the momentum integral with the pressure-
gradient term retained,
\[
\frac{d\theta}{dx} + (2 + H)\frac{\theta}{U_e}\frac{dU_e}{dx} =
\frac{C_f}{2},
\]
where $U_e(x)$ is the edge velocity from the inviscid outer flow and
$H = \delta^*/\theta$ the shape factor. Thwaites observed that the
dimensionless wall shear and the shape factor are nearly universal
functions of a single pressure-gradient parameter $\lambda =
(\theta^2/\nu)\,dU_e/dx$, and that the whole equation collapses to a
near-linear relation that integrates in closed form:
\[
\theta^2(x) = \frac{0.45\,\nu}{U_e^6(x)}\int_0^x U_e^5(\xi)\,d\xi.
\]
Once $\theta(x)$ is known, the correlated function $\lambda
\mapsto\{S(\lambda),H(\lambda)\}$ returns the wall shear and the
displacement thickness, and the local skin friction follows. The
device that makes Thwaites a separation predictor is its threshold:
the correlated wall-shear function reaches zero at $\lambda\approx
-0.09$, so the separation point is the streamwise station where the
computed $\lambda(x)$ falls to $-0.09$. A designer with only the
inviscid edge-velocity distribution over a body, from a panel method
or a potential-flow solution, can therefore march this one quadrature
along the surface, watch $\lambda$ fall through the adverse-gradient
region, and locate separation without solving the boundary-layer
partial differential equation at all. The method is accurate to a few
percent for attached laminar layers and is the standard hand-and-code
companion to the panel methods of the simulation exercises; its
turbulent counterpart is Head's entrainment method, which carries an
extra equation for the entrainment of freestream fluid into the
thicker turbulent layer.
\end{mastery}

\begin{mastery}
\textbf{The energy thickness and the Reynolds analogy.} The momentum
thickness is one of a family of integral thicknesses, each weighting
the velocity deficit by a different power of the local velocity. The
\engterm{energy thickness}
\[
\delta_e = \int_0^\infty \frac{u}{U_\infty}\left(1 - \frac{u^2}{
U_\infty^2}\right)dy
\]
measures the deficit in kinetic-energy flux the boundary layer
carries, and its transport equation, the mechanical-energy integral,
relates $d\delta_e/dx$ to the rate at which viscous stress dissipates
energy inside the layer. The same integral machinery applies to the
\engterm{thermal boundary layer}, where the temperature deficit
$(T - T_w)/(T_\infty - T_w)$ replaces the velocity ratio and a thermal
thickness plays the role of $\theta$. The momentum and thermal layers
are governed by equations of the same form, differing only in the
diffusivity: momentum diffuses at the kinematic viscosity $\nu$, heat
at the thermal diffusivity $\alpha$, and their ratio is the Prandtl
number $\text{Pr} = \nu/\alpha$. When $\text{Pr} = 1$ the two layers
are identical, which is the content of the \engterm{Reynolds analogy}:
the dimensionless wall heat flux, the Stanton number $\text{St}_H =
h/(\rho c_p U_\infty)$, equals half the skin-friction coefficient,
\[
\text{St}_H = \frac{C_f}{2} \quad (\text{Pr} = 1).
\]
For real fluids the Chilton-Colburn extension carries a Prandtl-number
correction, $\text{St}_H\,\text{Pr}^{2/3} = C_f/2$, accurate across
the engineering range of gases and many liquids. The analogy is the
working bridge between this chapter and the convective heat transfer
of Volume~IV: a skin-friction measurement or correlation predicts the
wall heat-transfer coefficient on the same surface without a separate
thermal calculation, which is why a single set of boundary-layer
relations serves both the drag engineer and the heat-transfer
engineer. The bridge breaks where the analogy's assumptions break,
under strong pressure gradients, separation, or large property
variation, exactly the regimes where the momentum layer itself
departs from its flat-plate behaviour.
\end{mastery}

\engterm{Turbulent boundary layer}. Above a critical Reynolds
number (typically $\text{Re}_x \sim 5\times 10^5$ for a smooth
flat plate, lower with surface roughness or freestream disturbance),
the laminar boundary layer transitions to turbulence. The turbulent
boundary layer is thicker than the laminar layer at the same
$\text{Re}_x$:
\[
\delta(x) \approx 0.37\,x\,\text{Re}_x^{-1/5}.
\]
The skin-friction coefficient is also larger:
\[
C_f(x) \approx 0.0592\,\text{Re}_x^{-1/5}, \quad
\bar{C}_f \approx 0.074\,\text{Re}_L^{-1/5}.
\]
The turbulent boundary layer has a different internal structure
than the laminar layer: a thin \engterm{viscous sublayer} near the
wall where $u^+ = y^+$ (with $u^+ = u/u_\tau$ the velocity scaled
by the friction velocity $u_\tau = \sqrt{\tau_w/\rho}$ and $y^+ =
y u_\tau/\nu$ the wall-distance in viscous units), a \engterm{log
layer} where $u^+ = (1/\kappa)\ln y^+ + B$ (with $\kappa \approx
0.41$, $B \approx 5.0$), and an \engterm{outer wake region} where
the profile deviates from the log law. This \engterm{law of the
wall} structure is one of the few quantitative universals of
turbulent flow; figure~\ref{fig:vol03ch12:law-of-wall} shows it in
the customary semilog form, where the viscous sublayer and the log
layer appear as a straight rising segment and a straight log
segment bridged by a buffer region. The structure also dictates
near-wall meshing in turbulence-resolving CFD: a wall-resolved
calculation needs the first grid point inside the viscous sublayer
($y^+\lesssim1$), while a wall-function calculation places it in
the log layer.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-law-of-the-wall}

\engterm{Pressure gradient effects}. A favourable pressure
gradient (pressure decreasing in the flow direction) thins the
boundary layer and stabilises it against transition; an adverse
pressure gradient (pressure increasing) thickens the layer and
can produce \engterm{separation}: the wall-shear stress reverses
sign and the flow detaches from the surface, forming a separated
wake. Separation is the canonical failure mode of boundary-layer
attached flow, and it underlies airfoil stall, diffuser stall,
bluff-body wake formation, and many of the high-drag
configurations the next section enumerates. The mechanism is drawn
in figure~\ref{fig:vol03ch12:separation}: the near-wall fluid
decelerates under the rising pressure, the wall shear stress
vanishes at the separation point, and the flow reverses and
detaches downstream.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-separation}

\section{Drag, lift, and the dimensionless numbers}\pathtag{standard}

The drag and lift on a body in a flow are conventionally
expressed in dimensionless form:
\[
C_D = \frac{F_D}{\tfrac{1}{2}\rho U^2 A}, \quad
C_L = \frac{F_L}{\tfrac{1}{2}\rho U^2 A},
\]
with $A$ a reference area (frontal area for drag, planform area
for lift). The coefficients depend on the body's shape and on the
dimensionless flow parameters. For incompressible viscous flow
without surface tension, the only dimensionless parameter is the
Reynolds number; for high-speed flow, the Mach number (chapter 13
of this volume) enters; for free-surface flow, the Froude number;
for surface-tension-dominated flow, the Weber number.

\engterm{Drag mechanisms}. The total drag on a body has two
components: \engterm{pressure drag} (form drag) from the integrated
pressure distribution over the body surface, and \engterm{skin-
friction drag} from the integrated viscous shear stress over the
body surface. The relative importance depends on the body shape
and on Reynolds number.

For a streamlined body (an airfoil at low angle of attack, a
fish-shaped fuselage), skin friction dominates: the pressure
recovers nearly fully over the rear of the body, and the drag
coefficient is small, $C_D \sim 0.01$ to $0.05$ for a well-designed
airfoil.

For a bluff body (a sphere, a cylinder, a cube), pressure drag
dominates: the flow separates from the rear of the body, leaving a
low-pressure wake that the surface pressure cannot recover. The
drag coefficient is large, $C_D \sim 0.4$ to $1.5$, and depends
weakly on Reynolds number above the transition regime.

For a flat plate at zero incidence (parallel to the flow), the
pressure drag vanishes and the drag coefficient equals the
skin-friction coefficient. For a flat plate normal to the flow,
the pressure drag dominates and $C_D \approx 2.0$ (the high end of
the bluff-body range).

\engterm{The drag crisis}. The drag coefficient of a sphere or
circular cylinder exhibits a characteristic drop near $\text{Re} =
3\times 10^5$, from $C_D \approx 0.5$ at slightly subcritical Re to
$C_D \approx 0.2$ at slightly supercritical Re. The drop is caused
by the laminar-turbulent transition in the boundary layer just
upstream of the separation point: a turbulent boundary layer
resists separation longer than a laminar one (because the turbulent
mixing transfers high-momentum freestream fluid to the near-wall
region), so the wake narrows and the pressure drag decreases
sharply. The drag crisis is the working basis of golf-ball dimple
design (the dimples trip the boundary layer to turbulence at lower
$\text{Re}$, lowering the drag in the operational range) and of
sandpaper strips on the front of competitive bicycle helmets
\cite{text:anderson-aerodynamics}. The drop is drawn for both a
sphere and a circular cylinder in
figure~\ref{fig:vol03ch12:drag-crisis}: each curve holds a
subcritical plateau, falls steeply through the crisis near
$\text{Re}=3\times10^5$, and recovers to a supercritical level
several times lower than the plateau.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-drag-crisis}

\engterm{The dimensionless numbers of fluid mechanics}. Beyond the
Reynolds number, several other dimensionless groups carry physical
meaning in specific regimes:

\begin{itemize}[leftmargin=*]
  \item \engterm{Froude number} $\text{Fr} = U/\sqrt{gL}$: ratio of
    inertial to gravitational forces. Governs free-surface flow
    (ship waves, river flow, weir discharge).
  \item \engterm{Mach number} $M = U/a$, with $a$ the local speed
    of sound: ratio of inertial to elastic-compressive forces.
    Governs compressible flow (chapter 13 of this volume).
  \item \engterm{Weber number} $\text{We} = \rho U^2 L/\sigma$, with
    $\sigma$ the surface tension: ratio of inertial to
    surface-tension forces. Governs droplet breakup, splashing,
    spray formation.
  \item \engterm{Knudsen number} $\text{Kn} = \lambda/L$, with
    $\lambda$ the molecular mean free path: ratio of molecular to
    continuum length scales. Continuum mechanics applies for
    $\text{Kn} < 10^{-2}$; rarefied gas dynamics for higher Kn.
  \item \engterm{Strouhal number} $\text{St} = fL/U$, with $f$ a
    characteristic oscillation frequency: governs unsteady flow,
    vortex shedding, flutter.
\end{itemize}

The working engineer keeps these in a short table and computes
the relevant ones before applying any flow analysis.

\engterm{The Kármán vortex street}. A circular cylinder in cross
flow at $\text{Re}$ between about $40$ and $200$ sheds vortices
alternately from its top and bottom surfaces, forming a regular
pattern of staggered vortices in the wake. The shedding frequency
is set by the Strouhal number: $\text{St} = fD/U \approx 0.21$ over
a wide range of $\text{Re}$. The shedding produces an oscillating
lift on the cylinder, which can drive resonant vibration of
slender structures: chimneys, suspension-bridge cables, transmission
lines, marine risers. The standard design discipline against vortex-
induced vibration is to ensure that the shedding frequency at the
operating wind speed does not coincide with the structure's
natural frequencies; aerodynamic devices (strakes, helical
fairings) on a cylinder disrupt the shedding pattern and reduce
the response amplitude.

\section{Turbulence: the unfinished problem}\pathtag{mastery}

Above a critical Reynolds number that depends on the geometry, a
laminar flow becomes unstable to small perturbations and
transitions to \engterm{turbulence}: a flow regime characterised
by chaotic, three-dimensional, multi-scale velocity fluctuations.
Turbulent flows have been the central unsolved problem of fluid
mechanics since Reynolds's 1883 paper
\cite{paper:reynolds1883}; the field has developed extensive
empirical, analytical, and computational tools, but a complete
predictive theory of turbulence at arbitrary Reynolds number does
not yet exist.

\engterm{Statistical description}. A turbulent velocity field is
decomposed into a mean and a fluctuation:
\[
u_i(\vec{x}, t) = \bar{u}_i(\vec{x}) + u'_i(\vec{x}, t),
\]
where the overbar denotes a time average. The Reynolds-averaged
Navier-Stokes (RANS) equation governs the mean field:
\[
\bar{u}_j\frac{\partial \bar{u}_i}{\partial x_j} = -\frac{1}{\rho}
\frac{\partial \bar{p}}{\partial x_i} + \nu\nabla^2 \bar{u}_i
- \frac{\partial (\overline{u'_i u'_j})}{\partial x_j}.
\]
The new term $-\overline{u'_i u'_j}$ is the \engterm{Reynolds
stress tensor}, a correlation of velocity fluctuations that acts
as an additional stress on the mean flow. The Reynolds stress is
not a closed-form function of the mean field; the \engterm{closure
problem} of turbulence is to model the Reynolds stress in terms of
mean-field quantities. No universally accurate closure exists; all
working models (mixing length, $k$-$\epsilon$, $k$-$\omega$,
Reynolds-stress models, large-eddy simulations) make assumptions
that are accurate in some flow regimes and inaccurate in others.

\begin{mastery}
\textbf{Turbulence closures, from mixing length to RANS models.}
The closure problem is to express the Reynolds stress
$-\overline{u'_i u'_j}$ in terms of the mean field. The oldest
device is Boussinesq's \engterm{eddy-viscosity hypothesis}: model
the Reynolds stress as if it were a viscous stress with an enhanced,
flow-dependent viscosity $\nu_t$,
\[
-\overline{u'_i u'_j} = \nu_t\!\left(\frac{\partial\bar u_i}{
\partial x_j} + \frac{\partial\bar u_j}{\partial x_i}\right)
- \tfrac{2}{3}k\,\delta_{ij},
\]
with $k = \tfrac12\overline{u'_i u'_i}$ the turbulent kinetic
energy. Prandtl's \engterm{mixing-length model} sets $\nu_t = \ell_m^2
|d\bar u/dy|$ with $\ell_m$ a geometry-dependent length; near a
wall $\ell_m = \kappa y$, which recovers the log law. Two-equation
models carry transport equations for two turbulence scalars: the
$k$-$\epsilon$ model solves for $k$ and the dissipation rate
$\epsilon$ and sets $\nu_t = C_\mu k^2/\epsilon$; the $k$-$\omega$
model solves for $k$ and a specific dissipation rate $\omega$ and
behaves better in adverse pressure gradients and near walls. The
shear-stress-transport (SST) model blends $k$-$\omega$ near the wall
with $k$-$\epsilon$ in the freestream. All of these are RANS models:
they predict the time-averaged field and are affordable enough for
routine design, but their accuracy depends on whether the calibration
flows match the flow at hand, and every one of them mispredicts
separation onset under strong adverse pressure gradients to some
degree. Resolving the eddies directly (LES, DNS) removes the closure
assumptions at large cost; the cost ratio between RANS and a
wall-resolved LES at engineering Reynolds numbers runs to several
orders of magnitude.
\end{mastery}

\engterm{The energy cascade}. Kolmogorov's 1941 theory of locally
isotropic turbulence holds that at sufficiently high Reynolds
number, energy injected at large scales (by some forcing) cascades
through intermediate scales toward small scales by inertial
nonlinear interactions, until the smallest scales (the
\engterm{Kolmogorov scale} $\eta = (\nu^3/\epsilon)^{1/4}$, with
$\epsilon$ the dissipation rate per unit mass) become small
enough that viscous dissipation converts the kinetic energy to
heat. In the inertial subrange between the forcing scale and the
Kolmogorov scale, the energy spectrum follows Kolmogorov's
$k^{-5/3}$ law:
\[
E(k) = C\epsilon^{2/3}k^{-5/3}, \quad C \approx 1.5.
\]
The Kolmogorov $-5/3$ scaling has been confirmed in many
turbulent flows at very high Reynolds number. The corresponding
ratio of largest to smallest length scales is $L/\eta \sim
\text{Re}^{3/4}$; resolving all scales of turbulence in a direct
numerical simulation (DNS) is therefore prohibitively expensive
for engineering Reynolds numbers (a flow at $\text{Re} = 10^6$
has $L/\eta \sim 30{,}000$, requiring $\sim 30{,}000^3 \sim
3\times 10^{13}$ grid points in three dimensions). DNS remains
restricted to research-scale Reynolds numbers; engineering
analysis relies on RANS or large-eddy simulation (LES) with
modelled closures.

\engterm{Practical consequences}. Turbulent flow in engineering
practice is recognised by:

\begin{itemize}[leftmargin=*]
  \item Higher skin-friction drag than laminar at the same Reynolds
    number (the turbulent boundary-layer scaling above).
  \item Higher heat transfer rate (the topic of Volume IV).
  \item Higher mass-transfer rate.
  \item Rapid mixing of dye, smoke, or other tracer scalars.
  \item Broadband acoustic radiation (the noise of a jet engine, a
    waterfall, a turbulent boundary layer on an aircraft wing).
  \item Resistance to separation under adverse pressure gradient
    (the basis of the drag crisis and of vortex generators on
    aircraft wings).
\end{itemize}

\section{Computational fluid dynamics introduced}\pathtag{mastery}

The Navier-Stokes equations admit no general analytical solution.
Numerical solution on a discretised grid is the working tool for
most engineering flow analysis; \engterm{computational fluid
dynamics} (CFD) is the body of techniques that does this.

The principal discretisation methods:

\begin{itemize}[leftmargin=*]
  \item \engterm{Finite difference}: derivatives are approximated
    on a structured grid. Simple to implement, accurate, but
    awkward on complex geometries.
  \item \engterm{Finite volume}: integral conservation laws are
    enforced on each grid cell. The standard method for industrial
    CFD; handles complex geometries through unstructured grids.
  \item \engterm{Finite element}: weak formulation of the
    governing equations, suitable for complex geometries and
    higher-order schemes; the standard method in structural
    mechanics and some CFD codes.
  \item \engterm{Spectral / spectral-element}: expansion in
    smooth basis functions, used for high-accuracy DNS and LES at
    moderate Reynolds numbers.
\end{itemize}

The choice between RANS, LES, and DNS depends on the Reynolds
number and the accuracy requirement:

\begin{itemize}[leftmargin=*]
  \item \engterm{RANS}: solves for the time-averaged flow with a
    turbulence closure model. Most affordable; appropriate for
    steady-state design calculations of bulk quantities (lift,
    drag, pressure distribution). Accuracy depends strongly on the
    closure model's match to the flow type.
  \item \engterm{LES}: resolves the large eddies and models the
    sub-grid scales. Computationally heavier than RANS by factors
    of $10^2$ to $10^4$; appropriate for unsteady flow physics
    where the large-scale dynamics dominates.
  \item \engterm{DNS}: resolves all scales. Computationally
    affordable only at modest Reynolds numbers ($\text{Re} \sim
    10^3$ to $10^5$ depending on geometry). Used for research and
    for validation of LES and RANS closures.
\end{itemize}

The discipline of CFD validation is to compare predictions
against experimental measurements at the same Reynolds number,
geometry, and boundary conditions. A CFD calculation that has
not been validated against experiment is a calculation, not a
prediction. The standard \engterm{verification and validation}
hierarchy distinguishes \engterm{verification} (the code solves
the equations correctly, checked by mesh refinement and
manufactured solutions) from \engterm{validation} (the equations
model the physics correctly, checked against experiment). Both
are required before a CFD prediction is treated as a design
basis.

\section{Worked examples}\pathtag{standard}

The relations of the preceding sections become a working method
only when they are run end to end on numbers. The worked examples
that follow each move from a stated configuration to a quantity an
engineer would report, and together they exercise every relation of
the chapter: the regime check and the friction factor in pipes and
ducts, the Blasius and turbulent boundary-layer scalings, the drag
laws across the Reynolds-number decades, the momentum integral under
a pressure gradient, the lubrication film, and the similarity rules
that scale a model to a prototype. The recurring discipline is the
one example~1 and example~3 establish: compute the Reynolds number
first, read the regime off it, and only then reach for the
regime-appropriate relation. An answer produced by the wrong-regime
formula is wrong by orders of magnitude, not by a correction term.

\subsection*{Worked example 1: regime, friction factor, and pressure drop in a water main}

Water at $15\,\degC$ ($\rho = 999\,\si{\kilogram\per\meter\cubed}$,
$\mu = 1.14\times10^{-3}\,\si{\pascal\second}$) flows at
$\bar u = 1.5\,\si{\meter\per\second}$ through a commercial-steel
main of diameter $D = 100\,\si{\milli\meter}$, length $L = 200\,\si{
\meter}$, absolute roughness $\epsilon = 0.045\,\si{\milli\meter}$.
Find the pressure drop.

The Reynolds number is
\[
\text{Re}_D = \frac{\rho \bar u D}{\mu} = \frac{999\times1.5\times
0.1}{1.14\times10^{-3}} \approx 1.31\times10^{5},
\]
well into the turbulent regime. The relative roughness is
$\epsilon/D = 0.045/100 = 4.5\times10^{-4}$. The Haaland correlation
gives
\[
\frac{1}{\sqrt{f}} = -1.8\log_{10}\!\left[
\left(\frac{4.5\times10^{-4}}{3.7}\right)^{1.11} +
\frac{6.9}{1.31\times10^{5}}\right]
= -1.8\log_{10}(3.6\times10^{-5} + 5.3\times10^{-5}) \approx 7.27,
\]
so $f \approx 1/7.27^2 \approx 0.0189$. The Darcy-Weisbach pressure
drop is
\[
\Delta p = f\frac{L}{D}\frac{\rho \bar u^2}{2} = 0.0189\times
\frac{200}{0.1}\times\frac{999\times1.5^2}{2} \approx 4.25\times10^4
\,\si{\pascal},
\]
about $42\,\text{kPa}$, or roughly $4.3\,\si{\meter}$ of water
head. A laminar estimate ($f = 64/\text{Re}_D \approx 4.9\times
10^{-4}$) would have under-predicted the drop by a factor near
$39$; computing the regime first is what keeps the answer right.

\subsection*{Worked example 2: laminar boundary layer on a flat plate}

Air at $20\,\degC$ ($\nu = 1.5\times10^{-5}\,\si{\meter}^2/\si{
\second}$, $\rho = 1.2\,\si{\kilogram\per\meter\cubed}$) flows at
$U_\infty = 3\,\si{\meter\per\second}$ over a plate of length
$L = 0.5\,\si{\meter}$. Find the boundary-layer thickness and the
total skin-friction drag per unit span.

The trailing-edge Reynolds number is
\[
\text{Re}_L = \frac{U_\infty L}{\nu} = \frac{3\times0.5}{1.5\times
10^{-5}} = 1.0\times10^{5},
\]
below the smooth-plate transition value near $5\times10^5$, so the
layer is laminar over the whole plate. The Blasius thickness at the
trailing edge is
\[
\delta(L) = \frac{5.0\,L}{\sqrt{\text{Re}_L}} = \frac{5.0\times0.5}{
\sqrt{1.0\times10^{5}}} \approx 7.9\times10^{-3}\,\si{\meter}
\approx 7.9\,\si{\milli\meter}.
\]
The average skin-friction coefficient is $\bar C_f = 1.328/\sqrt{
\text{Re}_L} = 1.328/316 \approx 4.2\times10^{-3}$, so the
drag per unit span (counting one side) is
\[
\frac{F_D}{b} = \tfrac12\rho U_\infty^2 L\,\bar C_f =
0.5\times1.2\times3^2\times0.5\times4.2\times10^{-3} \approx
1.1\times10^{-2}\,\si{\newton\per\meter}.
\]
The layer is millimetres thick and the drag is small; both are
characteristic of low-speed laminar flow over a short plate.

\subsection*{Worked example 3: terminal velocity of a settling particle}

A sand grain of diameter $d = 0.2\,\si{\milli\meter}$ and density
$\rho_s = 2650\,\si{\kilogram\per\meter\cubed}$ settles in water at
$20\,\degC$ ($\rho_f = 998\,\si{\kilogram\per\meter\cubed}$,
$\mu = 1.0\times10^{-3}\,\si{\pascal\second}$). Find the terminal
velocity.

Assume the Stokes regime and check afterward. Balancing weight minus
buoyancy against Stokes drag, $\tfrac{\pi}{6}d^3(\rho_s-\rho_f)g =
3\pi\mu d U$, gives
\[
U = \frac{(\rho_s-\rho_f)g\,d^2}{18\mu} = \frac{(2650-998)\times
9.81\times(2\times10^{-4})^2}{18\times1.0\times10^{-3}} \approx
3.6\times10^{-2}\,\si{\meter\per\second}.
\]
The check on the Reynolds number is $\text{Re} = \rho_f U d/\mu =
998\times0.036\times2\times10^{-4}/10^{-3} \approx 7.2$, which is
not below unity. The Stokes assumption is therefore marginal; the
Schiller-Naumann correlation ($C_D = (24/\text{Re})(1+0.15\,\text{
Re}^{0.687})$) lowers the answer to roughly $2.4\,\si{\centi\meter
\per\second}$, and the iterative solver in
\texttt{code/sphere\_terminal\_velocity.py} returns that value.
The lesson is the same as in example~1: the regime check is part of
the calculation, not an afterthought.

\subsection*{Worked example 4: vortex shedding off a chimney}

A steel chimney of diameter $D = 2\,\si{\meter}$ stands in a steady
wind of $U = 12\,\si{\meter\per\second}$ (air, $\nu = 1.5\times
10^{-5}\,\si{\meter}^2/\si{\second}$). Its lowest cross-wind natural
frequency is $f_n = 0.6\,\si{\hertz}$. Is vortex-induced resonance a
risk?

The Reynolds number is $\text{Re} = UD/\nu = 12\times2/(1.5\times
10^{-5}) = 1.6\times10^{6}$, in the range where the Strouhal number
holds near $\text{St}\approx0.2$. The shedding frequency is
\[
f_s = \text{St}\,\frac{U}{D} = 0.2\times\frac{12}{2} = 1.2\,\si{
\hertz}.
\]
The lock-in wind speed, where $f_s = f_n$, is $U_{\text{lock}} =
f_n D/\text{St} = 0.6\times2/0.2 = 6\,\si{\meter\per\second}$. A
$6\,\si{\meter\per\second}$ wind is a common condition, so the
chimney will pass through lock-in routinely; helical strakes or a
stiffening that raises $f_n$ are warranted.

\subsection*{Worked example 5: drag power on a road vehicle}

A car of frontal area $A = 2.2\,\si{\meter}^2$ and drag coefficient
$C_D = 0.30$ cruises at $U = 100\,\text{km/h} = 27.8\,\si{\meter\per
\second}$ in air ($\rho = 1.2\,\si{\kilogram\per\meter\cubed}$).
Find the drag force and the power spent against drag.

The drag force is
\[
F_D = \tfrac12\rho U^2 A\,C_D = 0.5\times1.2\times27.8^2\times2.2
\times0.30 \approx 3.1\times10^{2}\,\si{\newton},
\]
and the power is $P = F_D U \approx 306\times27.8 \approx 8.5\,\text{
kW}$. The cubic dependence on speed is the headline: raising the
cruise speed to $120\,\text{km/h}$ multiplies the drag power by
$(120/100)^3 \approx 1.73$, to about $14.7\,\text{kW}$. This is why
highway fuel economy falls off steeply with speed.

\subsection*{Worked example 6: scale-model similarity}

A $1/10$-scale model of a submarine is to be tested in a water tunnel
to predict the drag on the full vessel, which cruises at $U_p =
5\,\si{\meter\per\second}$. At what tunnel speed must the model run
for dynamic similarity, and what does the measured model drag imply
for the prototype?

Dynamic similarity in a single-phase, fully submerged, incompressible
flow requires matching the Reynolds number. With the same fluid
($\nu_m = \nu_p$) and the model length $L_m = L_p/10$,
\[
\frac{U_m L_m}{\nu} = \frac{U_p L_p}{\nu} \implies U_m = U_p
\frac{L_p}{L_m} = 5\times10 = 50\,\si{\meter\per\second}.
\]
A $50\,\si{\meter\per\second}$ water-tunnel speed is demanding, which
is one practical reason marine testing often accepts a Reynolds-number
mismatch and corrects for it. If similarity is achieved, the drag
coefficients match, $C_{D,m} = C_{D,p}$, so the prototype drag is
\[
F_{D,p} = F_{D,m}\,\frac{\rho_p U_p^2 A_p}{\rho_m U_m^2 A_m}.
\]
With $A_p = 100\,A_m$ and $U_p^2/U_m^2 = (5/50)^2 = 1/100$, the two
factors cancel and $F_{D,p} = F_{D,m}$: at matched Reynolds number
with the same fluid, the model and prototype carry the same drag
force, a result that surprises until the cancellation is seen.

\subsection*{Worked example 7: flat-plate drag from the momentum integral}

The Blasius drag coefficient $\bar C_f = 1.328/\sqrt{\text{Re}_L}$
can be recovered without the full similarity solution by feeding an
assumed velocity profile through the Kármán momentum integral. The
exercise shows how a one-parameter profile guess delivers an answer
within a few percent of the exact constant, which is the working
method for boundary layers that the Blasius solution does not cover.

Assume the cubic profile $u/U_\infty = \tfrac32(y/\delta) -
\tfrac12(y/\delta)^3$ for $0\le y\le\delta$ and $u = U_\infty$
beyond. The profile satisfies no-slip ($u=0$ at $y=0$), the
freestream match ($u=U_\infty$ at $y=\delta$), and zero curvature at
the wall ($\partial^2 u/\partial y^2 = 0$ at $y=0$, the wall
condition that follows from the momentum equation at zero pressure
gradient). The momentum thickness of this profile is
\[
\theta = \int_0^\delta \frac{u}{U_\infty}\!\left(1 -
\frac{u}{U_\infty}\right)dy = \frac{39}{280}\,\delta \approx
0.139\,\delta,
\]
and the wall shear is
\[
\tau_w = \mu\left(\frac{\partial u}{\partial y}\right)_{y=0} =
\frac{3\mu U_\infty}{2\delta}.
\]
The Kármán relation $d\theta/dx = \tau_w/(\rho U_\infty^2)$ becomes
\[
\frac{39}{280}\frac{d\delta}{dx} = \frac{3\mu}{2\rho U_\infty\,
\delta} \implies \delta\,d\delta = \frac{280}{39}\frac{3\nu}{2
U_\infty}\,dx.
\]
Integrating from $\delta=0$ at $x=0$ gives
\[
\delta(x) = \sqrt{\frac{280}{13}\frac{\nu x}{U_\infty}} \approx
4.64\,\sqrt{\frac{\nu x}{U_\infty}} = \frac{4.64\,x}{\sqrt{\text{Re}_x}},
\]
against the Blasius $5.0$, an error of about $7\,\%$ in the thickness.
The skin-friction coefficient follows from $\tau_w$:
\[
C_f(x) = \frac{\tau_w}{\tfrac12\rho U_\infty^2} = \frac{3\nu}{
U_\infty\,\delta} = \frac{0.646}{\sqrt{\text{Re}_x}},
\]
against the Blasius $0.664$, an error of about $2.7\,\%$, and the
length-averaged value is $\bar C_f = 1.292/\sqrt{\text{Re}_L}$
against the exact $1.328$. The lesson is that the momentum integral
turns a plausible profile into a drag estimate accurate to a few
percent, and the same machinery extends to pressure gradients and to
turbulent profiles where no Blasius solution exists.

\subsection*{Worked example 8: terminal velocity of a hailstone across drag regimes}

A spherical hailstone of diameter $d = 10\,\si{\milli\meter}$ and
density $\rho_s = 900\,\si{\kilogram\per\meter\cubed}$ falls through
air at sea level ($\rho_a = 1.2\,\si{\kilogram\per\meter\cubed}$,
$\mu = 1.8\times10^{-5}\,\si{\pascal\second}$). Find the terminal
velocity, carrying the regime check at every step.

A Stokes estimate ($U = (\rho_s-\rho_a)g d^2/(18\mu)$) gives
$U \approx 900\times9.81\times(10^{-2})^2/(18\times1.8\times10^{-5})
\approx 2.7\times10^{3}\,\si{\meter\per\second}$, which is absurd; the
Stokes Reynolds number it implies, $\text{Re} = \rho_a U d/\mu \sim
1.8\times10^6$, is nine orders of magnitude past the Stokes ceiling
of unity, so the Stokes law does not apply. The hailstone is a bluff
body in the Newton-plateau regime
(figure~\ref{fig:vol03ch12:terminal-velocity-regimes}), where
$C_D\approx0.44$ is roughly constant. Balance weight minus buoyancy
against the bluff-body drag,
\[
\tfrac{\pi}{6}d^3(\rho_s-\rho_a)g = \tfrac12\rho_a U^2\,
\tfrac{\pi}{4}d^2\,C_D,
\]
and solve for the velocity:
\[
U = \sqrt{\frac{4(\rho_s-\rho_a)g\,d}{3\rho_a C_D}} = \sqrt{\frac{
4\times900\times9.81\times10^{-2}}{3\times1.2\times0.44}} \approx
15\,\si{\meter\per\second}.
\]
The regime check confirms the choice: $\text{Re} = \rho_a U d/\mu =
1.2\times15\times10^{-2}/(1.8\times10^{-5}) \approx 1.0\times10^4$,
comfortably inside the $10^3$ to $2\times10^5$ Newton plateau where
$C_D\approx0.44$ holds. A $15\,\si{\meter\per\second}$ impact speed
(about $54\,\text{km/h}$) is consistent with the damage a centimetre
hailstone does to a car roof. The arithmetic is the same as the
raindrop estimate in the exercises, with the regime check deciding
which drag law enters the force balance.

\subsection*{Worked example 9: laminar entrance length in a fuel line}

Flow entering a pipe from a reservoir does not have its fully
developed profile at the inlet; the boundary layers grow from the
walls and merge over an \engterm{entrance length} $L_e$ before the
parabolic Hagen-Poiseuille profile is established. A fuel line
carries diesel ($\rho = 840\,\si{\kilogram\per\meter\cubed}$, $\mu =
3.0\times10^{-3}\,\si{\pascal\second}$) at $\bar u = 0.4\,\si{\meter
\per\second}$ through a $D = 8\,\si{\milli\meter}$ tube. Find the
entrance length and state whether a $0.5\,\si{\meter}$ tube run is
mostly developing or mostly developed.

The Reynolds number is
\[
\text{Re}_D = \frac{\rho\bar u D}{\mu} = \frac{840\times0.4\times
0.008}{3.0\times10^{-3}} \approx 896,
\]
below $2300$, so the flow is laminar and the laminar entrance-length
correlation $L_e \approx 0.05\,\text{Re}_D\,D$ applies:
\[
L_e \approx 0.05\times896\times0.008 \approx 0.36\,\si{\meter}.
\]
The entrance length is $0.36\,\si{\meter}$ out of a $0.5\,\si{\meter}$
run, so most of the tube is in developing flow, where the wall shear
and the pressure gradient are larger than the fully developed values.
A pressure-drop calculation that applied Hagen-Poiseuille over the
whole length would under-predict the drop, because the developing
region carries an extra pressure loss (the profile is being
accelerated near the centre and decelerated near the walls). The
correction is the entrance-loss term: for a laminar developing flow
the additional drop over the developing region is of order
$\rho\bar u^2$, comparable here to the fully developed drop, so it
cannot be neglected. Had the flow been turbulent, the entrance length
would have been far shorter ($L_e\approx4.4\,\text{Re}_D^{1/6}D$,
typically ten to forty diameters), which is why entrance effects are
a laminar-flow concern more than a turbulent one.

\subsection*{Worked example 10: displacement and momentum thickness from a given profile}

A boundary layer is approximated by the sinusoidal profile
$u/U_\infty = \sin(\pi y/2\delta)$ for $0\le y\le\delta$ and
$u = U_\infty$ beyond. Find the displacement thickness, the momentum
thickness, and the shape factor $H = \delta^*/\theta$, and compare to
the Blasius values. The three scales the profile carries are the ones
drawn in figure~\ref{fig:vol03ch12:thicknesses}.

The displacement thickness is the integral of the velocity deficit:
\[
\delta^* = \int_0^\delta\!\left(1 - \sin\frac{\pi y}{2\delta}\right)dy
= \delta - \left[-\frac{2\delta}{\pi}\cos\frac{\pi y}{2\delta}
\right]_0^\delta = \delta - \frac{2\delta}{\pi} = \delta\!\left(1 -
\frac{2}{\pi}\right) \approx 0.363\,\delta.
\]
The momentum thickness weights the deficit by the local velocity:
\[
\theta = \int_0^\delta \sin\frac{\pi y}{2\delta}\!\left(1 -
\sin\frac{\pi y}{2\delta}\right)dy.
\]
The first term integrates to $2\delta/\pi$; the second uses
$\int_0^\delta\sin^2(\pi y/2\delta)\,dy = \delta/2$, so
\[
\theta = \frac{2\delta}{\pi} - \frac{\delta}{2} = \delta\!\left(
\frac{2}{\pi} - \frac{1}{2}\right) \approx 0.137\,\delta.
\]
The shape factor is
\[
H = \frac{\delta^*}{\theta} = \frac{1 - 2/\pi}{2/\pi - 1/2} \approx
\frac{0.363}{0.137} \approx 2.66,
\]
against the Blasius value $H \approx 2.59$. The sinusoidal profile
recovers the displacement and momentum thicknesses and the shape
factor to within a few percent of the exact laminar values, which is
why the integral method built on an assumed profile (worked
example~7) lands so close to Blasius. The shape factor is itself a
working diagnostic: $H\approx2.6$ marks an attached laminar layer,
$H\approx1.4$ a turbulent one, and $H$ climbing toward $3.5$ to $4$
signals an adverse pressure gradient driving the layer toward
separation.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-thicknesses}

\subsection*{Worked example 11: Stokes drag and the relaxation time of an aerosol particle}

A spherical dust particle of diameter $d = 10\,\si{\micro\meter}$ and
density $\rho_p = 2000\,\si{\kilogram\per\meter\cubed}$ is carried in
air ($\rho_a = 1.2\,\si{\kilogram\per\meter\cubed}$, $\mu = 1.8\times
10^{-5}\,\si{\pascal\second}$). Find its still-air settling velocity,
its aerodynamic relaxation time, and the stopping distance it coasts
when the carrying air stops abruptly.

The settling velocity from the Stokes balance is
\[
U_s = \frac{(\rho_p - \rho_a)g\,d^2}{18\mu} = \frac{(2000-1.2)\times
9.81\times(10^{-5})^2}{18\times1.8\times10^{-5}} \approx 6.0\times
10^{-3}\,\si{\meter\per\second},
\]
about $6\,\si{\milli\meter\per\second}$. The Reynolds number check is
$\text{Re} = \rho_a U_s d/\mu = 1.2\times6.0\times10^{-3}\times10^{-5}
/(1.8\times10^{-5}) \approx 4\times10^{-3}\ll1$, so the Stokes law
applies and the answer stands. The aerodynamic relaxation time is the
ratio of the particle's inertia to the Stokes drag coefficient
$3\pi\mu d$:
\[
\tau_p = \frac{m}{3\pi\mu d} = \frac{\rho_p\,\tfrac{\pi}{6}d^3}{
3\pi\mu d} = \frac{\rho_p\,d^2}{18\mu} = \frac{2000\times(10^{-5})^2}{
18\times1.8\times10^{-5}} \approx 6.2\times10^{-4}\,\si{\second}.
\]
The relaxation time is the time the particle takes to forget its
initial velocity and follow the surrounding air (the same ratio that
gives $U_s = \tau_p g$, since the settling velocity is just the
relaxation time times gravity). The stopping distance, the length the
particle coasts when the air around it stops abruptly while the
particle moves at the air's former speed $U_0$, is
\[
\ell_{\text{stop}} = U_0\,\tau_p.
\]
For air moving at $U_0 = 1\,\si{\meter\per\second}$, the particle
coasts $\ell_{\text{stop}} = 1\times6.2\times10^{-4} \approx
0.62\,\si{\milli\meter}$ before matching the stopped air. The short
relaxation time and tiny stopping distance are why fine aerosol
particles track the air so faithfully that they make good flow
tracers, and why they are hard to capture by inertial impaction; the
$\tau_p\propto d^2$ scaling means a particle ten times larger has a
hundred times the relaxation time and is far easier to throw out of a
turning flow.

\subsection*{Worked example 12: skin-friction drag of a ship hull with laminar-turbulent transition}

A ship hull of waterline length $L = 150\,\si{\meter}$ and wetted
area $S = 4200\,\si{\meter\squared}$ moves at $U = 9\,\si{\meter\per
\second}$ (about $17.5$ knots) through seawater ($\rho = 1025\,\si{
\kilogram\per\meter\cubed}$, $\nu = 1.2\times10^{-6}\,\si{\meter}^2/
\si{\second}$). Estimate the skin-friction drag, treating the front
of the hull as laminar and the rest as turbulent, and compare to the
all-turbulent estimate.

The length Reynolds number is
\[
\text{Re}_L = \frac{U L}{\nu} = \frac{9\times150}{1.2\times10^{-6}}
\approx 1.1\times10^{9},
\]
enormous, so transition happens almost at the bow. Transition at
$\text{Re}_{x,\text{tr}}\approx5\times10^5$ occurs at
$x_{\text{tr}} = \nu\,\text{Re}_{x,\text{tr}}/U = 1.2\times10^{-6}
\times5\times10^5/9 \approx 0.067\,\si{\meter}$, under $7\,\si{\centi
\meter}$ from the bow on a $150\,\si{\meter}$ hull, so the laminar
patch is negligible and the all-turbulent estimate is the right one.
The all-turbulent average skin-friction coefficient is
\[
\bar C_f = \frac{0.074}{\text{Re}_L^{1/5}} = \frac{0.074}{(1.1\times
10^{9})^{0.2}} \approx \frac{0.074}{65} \approx 1.14\times10^{-3}.
\]
The skin-friction drag is
\[
F_D = \tfrac12\rho U^2 S\,\bar C_f = 0.5\times1025\times9^2\times4200
\times1.14\times10^{-3} \approx 2.0\times10^{5}\,\si{\newton}
\approx 200\,\text{kN}.
\]
The corresponding power against skin friction is $P = F_D U \approx
2.0\times10^5\times9 \approx 1.8\,\text{MW}$. A laminar-allowance
correction subtracts the small over-count of treating the first
$7\,\si{\centi\meter}$ as turbulent, which changes the answer by far
less than one percent, so the laminar patch is correctly ignored.
The lesson is the mirror image of worked example~9: at low Reynolds
number the developing or laminar region dominates, while at the huge
Reynolds number of a ship the layer is turbulent over nearly the
whole length, and the transition correction that matters on a model
hull at $\text{Re}_L\sim10^7$ is negligible on the full-scale ship.
This Reynolds-number gap is exactly why model-basin drag must be
scaled with care, splitting the measured drag into a friction part
that scales with $\text{Re}_L$ and a wave-making part that scales with
the Froude number, the procedure named for Froude.

\subsection*{Worked example 13: combined Couette-Poiseuille flow and the back-flow threshold}

The slot between two parallel plates a distance $h = 2\,\si{\milli
\meter}$ apart carries an oil ($\mu = 0.10\,\si{\pascal\second}$).
The upper plate at $y = h$ slides at $U = 0.30\,\si{\meter\per
\second}$ in the $+x$ direction; an adverse pressure gradient
$dp/dx = +1.2\times10^{4}\,\si{\pascal\per\meter}$ opposes the
drag flow. Find the velocity profile, the volume flow per unit
width, and the pressure gradient at which the flow at the
stationary wall first reverses.

The plane Navier-Stokes equation with no inertia reduces to
$\mu\,d^2u/dy^2 = dp/dx$, integrated twice with $u(0) = 0$ and
$u(h) = U$. The result is the superposition of the linear Couette
term and the parabolic pressure term,
\[
u(y) = U\frac{y}{h} - \frac{1}{2\mu}\frac{dp}{dx}\,y(h - y).
\]
Group the pressure term with the dimensionless number
$P = -\dfrac{h^2}{2\mu U}\dfrac{dp}{dx}$, which measures the
pressure forcing against the drag forcing. Here
\[
P = -\frac{(2\times10^{-3})^2}{2\times0.10\times0.30}\times
1.2\times10^{4} = -\frac{4\times10^{-6}}{0.06}\times1.2\times10^{4}
\approx -0.80,
\]
the negative sign marking an adverse gradient. The flow per unit
width is the integral of the profile,
\[
\frac{Q}{b} = \int_0^h u\,dy = \frac{Uh}{2} - \frac{h^3}{12\mu}
\frac{dp}{dx} = \frac{Uh}{2}\left(1 + \frac{P}{3}\right),
\]
which evaluates to
\[
\frac{Q}{b} = \frac{0.30\times2\times10^{-3}}{2}\left(1 +
\frac{-0.80}{3}\right) = 3.0\times10^{-4}\times0.733 \approx
2.2\times10^{-4}\,\si{\meter\squared\per\second}.
\]
The adverse gradient has cut the throughput to about three quarters
of the pure-Couette value. The wall shear at the stationary plate
is $\tau_0 = \mu(du/dy)_{y=0} = (\mu U/h)(1 + P)$, so the flow at
that wall reverses when $1 + P < 0$, that is $P < -1$. Setting
$P = -1$ gives the threshold pressure gradient
\[
\left.\frac{dp}{dx}\right|_{\text{rev}} = \frac{2\mu U}{h^2} =
\frac{2\times0.10\times0.30}{(2\times10^{-3})^2} = 1.5\times10^{4}
\,\si{\pascal\per\meter}.
\]
The applied $1.2\times10^{4}\,\si{\pascal\per\meter}$ sits below
this, so the profile is still everywhere forward, though the
near-wall slope is shallow. A gradient above $1.5\times10^{4}\,\si{
\pascal\per\meter}$ would drive a reverse-flow tongue along the
stationary wall, the parallel-plate analogue of boundary-layer
separation in section~12.5 and the flow-reversal sketched in
figure~\ref{fig:vol03ch12:couette-poiseuille}. The same algebra
governs the leakage flow in a hydraulic spool valve and the melt
flow in a single-screw extruder, where the screw provides the
drag term and the die back-pressure provides the adverse term.

\subsection*{Worked example 14: separation point from the Pohlhausen quartic profile}

The momentum-integral method of worked example~7 extends to a
pressure gradient through a one-parameter profile. Take the
Pohlhausen quartic
\[
\frac{u}{U_e} = 2\eta - 2\eta^3 + \eta^4 + \frac{\Lambda}{6}\,
\eta(1-\eta)^3, \quad \eta = \frac{y}{\delta},
\]
where $\Lambda = (\delta^2/\nu)\,dU_e/dx$ is the dimensionless
pressure-gradient parameter. Find the wall shear as a function of
$\Lambda$, the value of $\Lambda$ at which the layer separates, and
the shape factor there. The three members of this family are drawn
in figure~\ref{fig:vol03ch12:momentum-profiles}.

The wall shear follows from the wall slope of the profile:
\[
\tau_w = \mu\frac{U_e}{\delta}\left(\frac{du/U_e}{d\eta}\right)_{
\eta=0} = \mu\frac{U_e}{\delta}\left(2 + \frac{\Lambda}{6}\right).
\]
The wall shear vanishes, which is the definition of incipient
separation, when
\[
2 + \frac{\Lambda}{6} = 0 \implies \Lambda_{\text{sep}} = -12.
\]
A favourable gradient ($\Lambda > 0$) fills the profile and raises
the wall shear; the zero-gradient member $\Lambda = 0$ recovers the
flat-plate cubic-plus-quartic shape with wall slope $2$; the adverse
gradient $\Lambda = -12$ flattens the wall slope to zero. The
displacement and momentum thicknesses of the quartic are
\[
\frac{\delta^*}{\delta} = \frac{3}{10} - \frac{\Lambda}{120},
\quad
\frac{\theta}{\delta} = \frac{37}{315} - \frac{\Lambda}{945} -
\frac{\Lambda^2}{9072}.
\]
At separation, $\Lambda = -12$, these give $\delta^*/\delta =
0.3 + 0.1 = 0.40$ and $\theta/\delta = 0.1175 + 0.0127 - 0.0159
\approx 0.114$, so the shape factor is
\[
H = \frac{\delta^*}{\theta} = \frac{0.40}{0.114} \approx 3.5.
\]
The number is the working diagnostic worked example~10 anticipated:
an attached laminar layer carries $H\approx2.6$, and $H$ climbing
toward $3.5$ signals a layer on the edge of separation. The
Pohlhausen profile turns that qualitative warning into a computed
threshold, $\Lambda = -12$, that a designer marches along the
surface exactly as Thwaites' $\lambda = -0.09$ criterion is
marched; the two correlations differ in their constants because
they fit the wall shear with different assumed profiles, and both
land within a few percent of the exact Falkner-Skan separation
value.

\subsection*{Worked example 15: load capacity of a lubrication slider bearing}

A thrust pad $B = 50\,\si{\milli\meter}$ long (in the sliding
direction) and $W = 80\,\si{\milli\meter}$ wide rides over a runner
moving at $U = 5\,\si{\meter\per\second}$, separated by an oil film
($\mu = 2.0\times10^{-2}\,\si{\pascal\second}$) that converges from
$h_1 = 60\,\si{\micro\meter}$ at the inlet to $h_2 = 30\,\si{\micro
\meter}$ at the outlet. Find the load the film carries and the
friction power. The geometry and the pressure hump it generates are
drawn in figure~\ref{fig:vol03ch12:slider-bearing}.

The film is thin compared to its length, so the Reynolds lubrication
equation governs it; for a one-dimensional converging wedge the
integrated load per unit width is the standard result
\[
\frac{F}{W} = \frac{6\mu U B^2}{h_2^2}\,\frac{1}{(K-1)^2}\left[
\ln K - \frac{2(K-1)}{K+1}\right], \quad K = \frac{h_1}{h_2},
\]
where $K = 60/30 = 2$ is the convergence ratio. The bracket is
\[
\ln 2 - \frac{2(2-1)}{2+1} = 0.6931 - 0.6667 = 0.0265,
\]
and $1/(K-1)^2 = 1$, so
\[
\frac{F}{W} = \frac{6\times2.0\times10^{-2}\times5\times(0.050)^2}{
(30\times10^{-6})^2}\times0.0265 = \frac{1.5\times10^{-3}}{9.0\times
10^{-10}}\times0.0265 \approx 4.4\times10^{4}\,\si{\newton\per\meter}.
\]
Multiplying by the pad width $W = 0.080\,\si{\meter}$ gives a load
capacity $F \approx 3.5\times10^{3}\,\si{\newton}$, about
$3.5\,\text{kN}$ on a pad the size of a matchbox, which is the
working point of the slider: a thin converging oil film carries a
load several thousand times its own weight. The friction force on
the moving runner is the integral of the wall shear; for this wedge
it is
\[
\frac{F_f}{W} = \frac{\mu U B}{h_2}\,\frac{1}{K-1}\left[
4\ln K - \frac{6(K-1)}{K+1}\right] = \frac{2.0\times10^{-2}\times5
\times0.050}{30\times10^{-6}}\times(4\times0.6931 - 2) \approx
1.67\times10^{2}\times0.772 \approx 1.3\times10^{2}\,\si{\newton\per
\meter},
\]
so $F_f \approx 10\,\si{\newton}$ and the friction coefficient
referred to the load is $F_f/F \approx 10/3500 \approx 3\times
10^{-3}$, the same order as the journal bearing of the second case
study. The friction power is $P_f = F_f U \approx 10\times5 =
50\,\si{\watt}$, dissipated in the film and carried off by the oil.
The slider and the journal are the two canonical hydrodynamic
bearings, and both rest on the one fact this example makes concrete:
a viscous fluid dragged into a converging gap generates a pressure
that carries load without metal contact.

\subsection*{Worked example 16: turbulent flow in a non-circular duct by hydraulic diameter}

A rectangular HVAC duct of cross-section $a = 0.40\,\si{\meter}$ by
$b = 0.20\,\si{\meter}$ carries air ($\rho = 1.2\,\si{\kilogram\per
\meter\cubed}$, $\mu = 1.8\times10^{-5}\,\si{\pascal\second}$) at a
volume flow $Q = 0.50\,\si{\meter\cubed\per\second}$ over a run
$L = 30\,\si{\meter}$ of galvanised steel ($\epsilon = 0.15\,\si{
\milli\meter}$). Find the pressure drop, using the hydraulic
diameter to carry the circular-pipe machinery over to the
rectangular section.

The mean velocity is
\[
\bar u = \frac{Q}{ab} = \frac{0.50}{0.40\times0.20} = 6.25\,\si{
\meter\per\second}.
\]
The hydraulic diameter is four times the area over the wetted
perimeter,
\[
D_h = \frac{4A}{P} = \frac{4ab}{2(a+b)} = \frac{4\times0.40\times
0.20}{2\times0.60} = 0.267\,\si{\meter}.
\]
The hydraulic-diameter Reynolds number is
\[
\text{Re}_{D_h} = \frac{\rho\bar u D_h}{\mu} = \frac{1.2\times6.25
\times0.267}{1.8\times10^{-5}} \approx 1.1\times10^{5},
\]
turbulent. The relative roughness is $\epsilon/D_h = 0.15/267 =
5.6\times10^{-4}$, and the Haaland correlation gives
\[
\frac{1}{\sqrt f} = -1.8\log_{10}\!\left[\left(\frac{5.6\times
10^{-4}}{3.7}\right)^{1.11} + \frac{6.9}{1.1\times10^{5}}\right]
= -1.8\log_{10}(4.8\times10^{-5} + 6.3\times10^{-5}) \approx 7.10,
\]
so $f \approx 1/7.10^2 \approx 0.0198$. The Darcy-Weisbach drop,
with $D_h$ in place of the circular diameter, is
\[
\Delta p = f\frac{L}{D_h}\frac{\rho\bar u^2}{2} = 0.0198\times
\frac{30}{0.267}\times\frac{1.2\times6.25^2}{2} \approx 2.22\times
23.4 \approx 52\,\si{\pascal}.
\]
A $52\,\si{\pascal}$ drop over a $30\,\si{\meter}$ run is a typical
duct loss, the kind a fan must overcome at a few tens of pascals per
straight run plus the fitting losses at every bend and transition.
The hydraulic-diameter substitution is exact only for the friction
factor's leading behaviour; the laminar constant changes with
aspect ratio (it is $64$ for a circle but $56.9$ for this $2{:}1$
rectangle and $96$ for infinite parallel plates), so the device is
a turbulent-flow approximation good to a few percent, and a precise
laminar duct calculation uses the aspect-ratio-specific constant
instead.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-momentum-profiles}

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-slider-bearing}

\subsection*{Worked example 17: wind-tunnel similarity for a car at reduced scale}

A $1/4$-scale model of a car is tested in a wind tunnel to predict the
full-scale drag at a road speed of $U_p = 30\,\si{\meter\per\second}$
(about $108\,\text{km/h}$). The model and the full car run in the same
air at sea level ($\rho = 1.2\,\si{\kilogram\per\meter\cubed}$, $\nu =
1.5\times10^{-5}\,\si{\meter}^2/\si{\second}$). Find the tunnel speed
that matches Reynolds number, judge whether it is attainable, and state
how a model drag-area measurement transfers to the prototype.

Dynamic similarity for this single-phase incompressible external flow
requires matching the Reynolds number. With the same fluid and the
model length $L_m = L_p/4$,
\[
\frac{U_m L_m}{\nu} = \frac{U_p L_p}{\nu} \implies U_m = U_p
\frac{L_p}{L_m} = 30\times4 = 120\,\si{\meter\per\second}.
\]
A $120\,\si{\meter\per\second}$ tunnel speed is about Mach $0.35$, high
enough that compressibility begins to intrude (the incompressible
assumption frays above roughly Mach $0.3$), and few low-speed tunnels
reach it. The practical resolutions are to accept the Reynolds-number
mismatch and trip the model boundary layer to turbulence with roughness
strips so that it separates as the full-scale flow does, or to use a
pressurised tunnel that raises $\rho$ and so raises the model Reynolds
number at a modest speed. Suppose similarity is achieved well enough
that the drag area transfers. The model is measured to carry a drag
$F_{D,m} = 95\,\si{\newton}$ at the matched condition. The drag
coefficient is the same on model and prototype, so the drag area
$C_D A$ scales with the measured force through the dynamic pressure and
the area. Equating drag coefficients,
\[
\frac{F_{D,m}}{\tfrac12\rho U_m^2 A_m} = \frac{F_{D,p}}{\tfrac12\rho
U_p^2 A_p},
\]
and with $A_p = 16\,A_m$ and $U_p^2/U_m^2 = (30/120)^2 = 1/16$, the
prototype force is
\[
F_{D,p} = F_{D,m}\,\frac{U_p^2}{U_m^2}\,\frac{A_p}{A_m} = 95\times
\frac{1}{16}\times16 = 95\,\si{\newton}.
\]
The two scale factors cancel exactly, as they did for the submarine of
worked example~6: at matched Reynolds number in the same fluid the
model and prototype carry the same drag force, which is the
dimensional-analysis result of section~12.4 made concrete. The lesson
the cancellation hides is that the equality holds only if the Reynolds
numbers match; a model run at the same speed as the prototype, at
one-quarter the Reynolds number, would read a drag coefficient off the
wrong part of the $C_D(\text{Re})$ curve, the error worked example~19
below quantifies for a cylinder near the drag crisis.

\subsection*{Worked example 18: plane Poiseuille flow and the flow-rate law}

Glycerin-water at $\mu = 0.20\,\si{\pascal\second}$ is driven between
two long stationary parallel plates a gap $h = 3\,\si{\milli\meter}$
apart and $w = 0.25\,\si{\meter}$ wide by a pressure gradient
$dp/dx = -8.0\times10^{3}\,\si{\pascal\per\meter}$. Derive the
velocity profile and the flow rate, and find the wall shear and the
ratio of mean to maximum velocity.

With no wall motion and no inertia the plane Navier-Stokes equation is
$\mu\,d^2u/dy^2 = dp/dx$, a constant on the right. Integrate twice,
$\mu\,u = \tfrac12(dp/dx)y^2 + C_1 y + C_2$, and apply no-slip at both
walls, $u(0) = u(h) = 0$. The conditions give $C_2 = 0$ and
$C_1 = -\tfrac12(dp/dx)h$, so
\[
u(y) = -\frac{1}{2\mu}\frac{dp}{dx}\,y(h - y),
\]
a parabola symmetric about the mid-gap, the planar counterpart of the
Hagen-Poiseuille pipe parabola. The maximum is on the centreline
$y = h/2$,
\[
u_{\max} = -\frac{1}{8\mu}\frac{dp}{dx}\,h^2 = \frac{8.0\times10^{3}
\times(3\times10^{-3})^2}{8\times0.20} = \frac{8.0\times10^{3}\times
9\times10^{-6}}{1.6} \approx 4.5\times10^{-2}\,\si{\meter\per\second}.
\]
The flow rate per unit width is the integral of the profile,
\[
\frac{Q}{w} = \int_0^h u\,dy = -\frac{1}{12\mu}\frac{dp}{dx}\,h^3 =
\frac{8.0\times10^{3}\times(3\times10^{-3})^3}{12\times0.20} \approx
9.0\times10^{-5}\,\si{\meter\squared\per\second},
\]
so the total flow is $Q = (Q/w)\,w = 9.0\times10^{-5}\times0.25 \approx
2.3\times10^{-5}\,\si{\meter\cubed\per\second}$, about
$1.4\,\text{L/min}$. The mean velocity is $\bar u = (Q/w)/h = 9.0\times
10^{-5}/(3\times10^{-3}) = 3.0\times10^{-2}\,\si{\meter\per\second}$,
exactly two-thirds of the maximum, the planar value against the
pipe's one-half. The wall shear is the wall slope,
\[
\tau_w = \mu\left(\frac{du}{dy}\right)_{y=0} = -\frac12\frac{dp}{dx}\,h
= \frac12\times8.0\times10^{3}\times3\times10^{-3} = 12\,\si{\pascal}.
\]
The Reynolds-number check, on the gap as length scale with the fluid
density $\rho\approx1200\,\si{\kilogram\per\meter\cubed}$, is
$\text{Re} = \rho\bar u h/\mu = 1200\times0.030\times0.003/0.20\approx
0.54$, deep in the laminar viscous regime, so the no-inertia derivation
holds. The $Q\propto h^3$ scaling is the planar analogue of the pipe's
$Q\propto D^4$: a gap halved cuts the throughput eightfold, which is
why coating-die and extrusion-slot gaps are held to tight tolerance.

\subsection*{Worked example 19: drag of a cylinder across the drag crisis}

A circular cylinder (a bridge hanger) of diameter $D = 0.20\,\si{
\meter}$ stands in a cross wind. Find the drag force per unit length at
$U_1 = 15\,\si{\meter\per\second}$ and at $U_2 = 40\,\si{\meter\per
\second}$ (air, $\nu = 1.5\times10^{-5}\,\si{\meter}^2/\si{\second}$,
$\rho = 1.2\,\si{\kilogram\per\meter\cubed}$), and show how the drag
crisis makes the higher wind speed carry barely more drag than the
lower.

The Reynolds numbers are
\[
\text{Re}_1 = \frac{U_1 D}{\nu} = \frac{15\times0.20}{1.5\times10^{-5}}
= 2.0\times10^{5}, \qquad
\text{Re}_2 = \frac{U_2 D}{\nu} = \frac{40\times0.20}{1.5\times10^{-5}}
\approx 5.3\times10^{5}.
\]
The first sits on the subcritical plateau, where the cylinder drag
coefficient is $C_{D,1}\approx1.2$; the second sits above the drag
crisis near $\text{Re} = 3\times10^5$
(figure~\ref{fig:vol03ch12:drag-crisis}), where the coefficient has
fallen to $C_{D,2}\approx0.4$. The drag per unit length is $\tfrac12
\rho U^2 D\,C_D$. At the lower speed,
\[
\frac{F_1}{\ell} = \tfrac12\times1.2\times15^2\times0.20\times1.2
\approx 32\,\si{\newton\per\meter},
\]
and at the higher speed,
\[
\frac{F_2}{\ell} = \tfrac12\times1.2\times40^2\times0.20\times0.4
\approx 77\,\si{\newton\per\meter}.
\]
The wind speed rose by a factor of $2.7$, which on a constant drag
coefficient would have raised the force by $2.7^2\approx7.1$ to about
$230\,\si{\newton\per\meter}$; the drag-crisis drop in $C_D$ from
$1.2$ to $0.4$ pulls the actual rise down to a factor of $2.4$. The
hazard the example warns of is the design that reads the wrong side of
the crisis: a wind-tunnel test of a smooth model in a quiet tunnel may
sit subcritical and report $C_D\approx1.2$ at a Reynolds number where
the rough, turbulence-bathed full-scale cylinder runs supercritical at
$C_D\approx0.4$, a factor-of-three over-estimate of the design wind
load. The regime check here is a reading of which side of the crisis
the operating Reynolds number falls on, the bluff-body counterpart of
the laminar-turbulent regime check that opened the worked examples.

\subsection*{Worked example 20: viscous diffusion depth in an oscillating flow}

A flat plate at the bottom of a tank of oil ($\nu = 1.0\times10^{-4}\,
\si{\meter}^2/\si{\second}$) oscillates in its own plane at angular
frequency $\omega = 2\pi\,\si{\radian\per\second}$ (one cycle per
second). Find the depth over which the oscillation is felt in the
fluid, the Stokes-layer thickness, and compare it to the steady viscous
diffusion depth after one second.

The plate drives Stokes's second problem: an oscillating wall velocity
$U\cos\omega t$ launches a damped shear wave into the fluid whose
amplitude decays as $\exp(-y/\delta_s)$ with the \engterm{Stokes-layer
thickness}
\[
\delta_s = \sqrt{\frac{2\nu}{\omega}} = \sqrt{\frac{2\times1.0\times
10^{-4}}{2\pi}} \approx \sqrt{3.18\times10^{-5}} \approx 5.6\times
10^{-3}\,\si{\meter},
\]
about $5.6\,\si{\milli\meter}$. The oscillation is effectively confined
to a layer a few times $\delta_s$, so a centimetre or so of oil above
the plate feels the motion and the bulk above stays still. The
thickness shrinks as $\omega^{-1/2}$: a plate oscillating a hundred
times faster drives a Stokes layer ten times thinner, which is why
high-frequency vibration is felt only in a very thin fluid film and why
an acoustic boundary layer is microns thick. Compare the steady viscous
diffusion depth of Stokes's first problem after one second, $\delta
\sim \sqrt{\nu t} = \sqrt{1.0\times10^{-4}\times1} = 1.0\times10^{-2}
\,\si{\meter}$, about $1\,\si{\centi\meter}$. The two depths are the
same order because one second is roughly one oscillation period; the
distinction is that the steady depth keeps growing as $\sqrt{t}$ while
the oscillating depth is pinned at $\delta_s$, because each half-cycle
reverses the diffusion the previous one drove. The $\sqrt{\nu/\omega}$
scale recurs wherever a fluid meets a vibrating surface: the acoustic
streaming near a loudspeaker cone, the secondary flow in a pulsatile
artery, and the damping of a quartz-crystal microbalance in a liquid.

\subsection*{Worked example 21: friction velocity and near-wall mesh spacing for CFD}

A wall-resolved CFD calculation of air flow ($\rho = 1.2\,\si{\kilogram
\per\meter\cubed}$, $\nu = 1.5\times10^{-5}\,\si{\meter}^2/\si{\second}$)
over a flat surface at $U_\infty = 30\,\si{\meter\per\second}$ must
place its first grid point inside the viscous sublayer, at $y^+ \le 1$,
at a streamwise station where $\text{Re}_x = 5\times10^{6}$. Find the
wall shear, the friction velocity, and the height of the first cell.

The turbulent local skin-friction coefficient at this station is
\[
C_f = \frac{0.0592}{\text{Re}_x^{1/5}} = \frac{0.0592}{(5\times10^{6})
^{0.2}} \approx \frac{0.0592}{21.9} \approx 2.7\times10^{-3}.
\]
The wall shear stress follows from the definition of $C_f$,
\[
\tau_w = C_f\,\tfrac12\rho U_\infty^2 = 2.7\times10^{-3}\times0.5\times
1.2\times30^2 \approx 1.46\,\si{\pascal},
\]
and the \engterm{friction velocity} is
\[
u_\tau = \sqrt{\frac{\tau_w}{\rho}} = \sqrt{\frac{1.46}{1.2}} \approx
1.10\,\si{\meter\per\second},
\]
about $3.7\,\%$ of the freestream, the usual order for a turbulent
boundary layer. The wall-unit length is $\nu/u_\tau$, so the cell
height that places the first node at $y^+ = 1$ is
\[
y_1 = \frac{y^+\,\nu}{u_\tau} = \frac{1\times1.5\times10^{-5}}{1.10}
\approx 1.4\times10^{-5}\,\si{\meter} \approx 14\,\si{\micro\meter}.
\]
A first cell of about fourteen microns against a boundary layer that is
centimetres thick is the price of resolving the wall: the near-wall
spacing must be three to four orders of magnitude finer than the layer
it sits in, which is why wall-resolved meshes carry so many cells and
why wall-function models, which place the first node in the log layer
at $y^+\approx30$ to $300$ and bridge the sublayer with the law of the
wall, are the affordable alternative. Scaling the same calculation to a
faster flow makes it worse: $u_\tau$ grows with $U_\infty$, so the
required first-cell height falls and the cell count climbs, the
quantitative form of the near-wall meshing remark in section~12.4 on
the law of the wall.

\subsection*{Applied vignette: the falling-sphere viscometer}

The Stokes drag law of section~12.4 is the working principle of one of
the oldest instruments for measuring viscosity. A small sphere is
dropped into a vertical tube of the test fluid and timed as it falls at
its terminal velocity between two marks a distance $L$ apart
(figure~\ref{fig:vol03ch12:falling-sphere}). At terminal velocity the
sphere is in equilibrium, so the downward weight balances the upward
sum of buoyancy and Stokes drag,
\[
\tfrac{4}{3}\pi a^3\rho_s\,g = \tfrac{4}{3}\pi a^3\rho_f\,g + 6\pi\mu
a U,
\]
where $a$ is the sphere radius, $\rho_s$ its density, $\rho_f$ the
fluid density, and $U = L/t$ the measured terminal velocity. Solving
for the viscosity,
\[
\mu = \frac{2 a^2(\rho_s - \rho_f)\,g}{9 U} = \frac{2 a^2(\rho_s -
\rho_f)\,g\,t}{9 L}.
\]
The measurement is a stopwatch reading: every other quantity is fixed
by the sphere and the fluid, so timing the fall returns the viscosity
directly. Take a steel ball of radius $a = 1.0\,\si{\milli\meter}$
($\rho_s = 7800\,\si{\kilogram\per\meter\cubed}$) falling in a glycerin
sample ($\rho_f = 1260\,\si{\kilogram\per\meter\cubed}$), timed over
$L = 0.20\,\si{\meter}$ at $t = 2.5\,\si{\second}$, so $U = 0.080\,\si{
\meter\per\second}$. The viscosity reads
\[
\mu = \frac{2\times(10^{-3})^2\times(7800 - 1260)\times9.81}{9\times
0.080} \approx \frac{2\times10^{-6}\times6540\times9.81}{0.72} \approx
0.18\,\si{\pascal\second},
\]
in the right range for a glycerin-water blend at room temperature. The
reading is trustworthy only inside the Stokes regime, so the
measurement carries its own regime check: the sphere Reynolds number is
\[
\text{Re} = \frac{\rho_f U (2a)}{\mu} = \frac{1260\times0.080\times
2\times10^{-3}}{0.18} \approx 1.1,
\]
just at the Stokes ceiling, so the simple law is marginal and a precise
instrument either uses a smaller or lighter ball to drop the Reynolds
number well below unity or applies the Oseen inertial correction
$C_D = (24/\text{Re})(1 + \tfrac{3}{16}\text{Re})$ to the drag. A
second correction matters in a real tube: the finite tube wall retards
the sphere, and the measured velocity is multiplied by a wall factor
$(1 - 2.10\,a/R_{\text{tube}})$ to first order in the diameter ratio
(the Faxen correction), so a commercial falling-sphere viscometer fixes
the ball-to-tube ratio and tabulates the factor. With both corrections
the instrument is accurate to better than one percent over the
viscosity decade from light oils to glycerin, which is why it remained
a laboratory standard for transparent Newtonian liquids long after
rotational viscometers arrived. The same physics, read backward, is the
particle-settling calculation of worked examples~3 and~11: the
viscometer fixes the fluid and infers the sphere's speed, while the
settling problem fixes the sphere and infers its speed in a known
fluid.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-falling-sphere}

\section{End-to-end case study: sizing a pump for a transfer line}\pathtag{standard}

The relations of this chapter become a design method when they are
run together on one system from the fluid and the geometry through to
a sized pump and a verdict. We size a centrifugal pump to move water
from a supply sump to an elevated process tank, working the chain
Reynolds number $\to$ friction factor $\to$ major loss $\to$ minor
losses $\to$ system curve $\to$ operating point $\to$ required power
$\to$ cavitation margin, carrying the numbers at every link. The
layout is the one drawn in figure~\ref{fig:vol03ch12:pump-layout}.

\paragraph{The duty and the fluid.} The plant requires a delivery of
$Q = 25\,\si{\liter\per\second} = 0.025\,\si{\meter\cubed\per\second}$
of water at $20\,\degC$ ($\rho = 998\,\si{\kilogram\per\meter\cubed}$,
$\mu = 1.0\times10^{-3}\,\si{\pascal\second}$, vapour pressure
$p_v = 2.34\,\si{\kilo\pascal}$ absolute). The supply sump free
surface sits $2\,\si{\meter}$ below the pump centreline (a suction
lift), and the delivery tank free surface sits $18\,\si{\meter}$
above the sump surface, so the static lift the pump must overcome is
$H_s = 18\,\si{\meter}$. The pipe is commercial steel of internal
diameter $D = 150\,\si{\milli\meter}$ and roughness $\epsilon =
0.045\,\si{\milli\meter}$; the suction run is $6\,\si{\meter}$ and the
discharge run is $120\,\si{\meter}$, for a total straight length
$L = 126\,\si{\meter}$.

\paragraph{Velocity and Reynolds number.} The bore area is
\[
A = \frac{\pi D^2}{4} = \frac{\pi (0.150)^2}{4} = 1.767\times10^{-2}
\,\si{\meter\squared},
\]
so the mean velocity is
\[
\bar u = \frac{Q}{A} = \frac{0.025}{1.767\times10^{-2}} \approx
1.41\,\si{\meter\per\second},
\]
a sensible duty velocity (the usual design band for water in steel is
$1$ to $3\,\si{\meter\per\second}$). The Reynolds number is
\[
\text{Re}_D = \frac{\rho\bar u D}{\mu} = \frac{998\times1.41\times
0.150}{1.0\times10^{-3}} \approx 2.1\times10^{5},
\]
fully turbulent.

\paragraph{Friction factor from Colebrook, checked against Haaland.}
The relative roughness is $\epsilon/D = 0.045/150 = 3.0\times10^{-4}$.
The implicit Colebrook-White equation
\[
\frac{1}{\sqrt f} = -2\log_{10}\!\left(\frac{\epsilon/D}{3.7} +
\frac{2.51}{\text{Re}_D\sqrt f}\right)
\]
is solved by iteration. Starting from a guess $f_0 = 0.02$
($\sqrt{f_0}=0.1414$): the right side is $-2\log_{10}(8.1\times10^{-5}
+ 2.51/(2.1\times10^5\times0.1414)) = -2\log_{10}(8.1\times10^{-5} +
8.45\times10^{-5}) = -2\log_{10}(1.66\times10^{-4}) = 7.56$, giving
$f_1 = 1/7.56^2 = 0.0175$. One more pass with $\sqrt{f_1}=0.1323$
gives $-2\log_{10}(8.1\times10^{-5} + 9.03\times10^{-5}) =
-2\log_{10}(1.71\times10^{-4}) = 7.53$, so $f = 1/7.53^2 \approx
0.0176$; the iteration has converged. The explicit Haaland form
returns $f\approx0.0177$, a $0.6\,\%$ difference, which confirms the
root. Take $f = 0.0176$. The solver behind
figure~\ref{fig:vol03ch12:moody} carries both forms in
\texttt{code/moody.py}.

\paragraph{Major loss along the pipe.} The Darcy-Weisbach head loss
over the straight run is
\[
h_{f,\text{major}} = f\frac{L}{D}\frac{\bar u^2}{2g} = 0.0176\times
\frac{126}{0.150}\times\frac{1.41^2}{2\times9.81} \approx 14.79\times
0.101 \approx 1.50\,\si{\meter}.
\]
(Working in head divides the Darcy-Weisbach pressure drop by $\rho g$;
the friction head is $1.50\,\si{\meter}$ of water, a pressure drop of
$\rho g h_f \approx 14.7\,\si{\kilo\pascal}$.)

\paragraph{Minor losses from fittings.} The line carries a
square-edged entrance ($K=0.5$), four $90^{\circ}$ long-radius elbows
($K=0.3$ each), one fully open gate valve ($K=0.15$), one swing check
valve ($K=2.0$), and an exit into the tank ($K=1.0$). The summed loss
coefficient is
\[
\sum K = 0.5 + 4\times0.3 + 0.15 + 2.0 + 1.0 = 4.85,
\]
and the minor-loss head is
\[
h_{f,\text{minor}} = \left(\sum K\right)\frac{\bar u^2}{2g} = 4.85
\times\frac{1.41^2}{2\times9.81} \approx 4.85\times0.101 \approx
0.49\,\si{\meter}.
\]
The fitting losses are about a third of the pipe-friction loss here,
which is typical for a run of this length; on a short, fitting-heavy
skid the minor losses can dominate, and on a long cross-country line
they are negligible. The total friction head at the duty flow is
$h_f = 1.50 + 0.49 \approx 1.99\,\si{\meter}$.

\paragraph{System curve and operating point.} The head the pump must
supply at flow $Q$ is the static lift plus the friction head, and the
friction head scales as $\bar u^2 \propto Q^2$:
\[
H_{\text{sys}}(Q) = H_s + \left(f\frac{L}{D} + \sum K\right)
\frac{\bar u^2}{2g} = 18 + k\,Q^2,
\]
where, with $\bar u = Q/A$, the coefficient is
\[
k = \left(f\frac{L}{D} + \sum K\right)\frac{1}{2gA^2} = (14.79 +
4.85)\frac{1}{2\times9.81\times(1.767\times10^{-2})^2} \approx 3204\,
\si{\meter\per(\meter\cubed\per\second)\squared}.
\]
At the duty flow $Q = 0.025\,\si{\meter\cubed\per\second}$ this gives
$H_{\text{sys}} = 18 + 3204\times0.025^2 = 18 + 2.00 = 20.0\,\si{
\meter}$, consistent with the static lift plus the $1.99\,\si{\meter}$
friction head computed above. A candidate pump has the falling
characteristic $H_{\text{pump}}(Q) = 55 - b\,Q^2$ with $b \approx
1.97\times10^4$ in the same units (a typical single-stage end-suction
curve; the manufacturer supplies it as a chart, here reduced to a
parabola for the algebra). The operating point is where supply meets
demand, $H_{\text{pump}} = H_{\text{sys}}$:
\[
55 - 1.97\times10^4\,Q^2 = 18 + 3204\,Q^2 \implies Q^2 = \frac{37}{
2.29\times10^4} \implies Q \approx 0.0402\,\si{\meter\cubed\per
\second}.
\]
The unthrottled pump would run at about $40\,\si{\liter\per\second}$,
above the $25\,\si{\liter\per\second}$ duty, so the operating point is
trimmed back to the duty flow by closing a control valve (raising
$\sum K$ until the system curve steepens to cross the pump curve at
$25\,\si{\liter\per\second}$) or, more efficiently, by reducing the
impeller speed. The crossing is drawn in
figure~\ref{fig:vol03ch12:pump-system-curve}. At the duty flow the
pump head is read from the curve as $H = 20\,\si{\meter}$ once it is
trimmed to the duty point.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-pump-layout}

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-pump-system-curve}

\paragraph{Required power.} The hydraulic power delivered to the fluid
at the duty point is
\[
P_{\text{hyd}} = \rho g Q H = 998\times9.81\times0.025\times20.0
\approx 4.90\times10^{3}\,\si{\watt} \approx 4.9\,\si{\kilo\watt}.
\]
A single-stage water pump of this size runs at an efficiency near
$\eta\approx0.75$, so the shaft power is $P_{\text{shaft}} =
P_{\text{hyd}}/\eta \approx 4.9/0.75 \approx 6.5\,\si{\kilo\watt}$,
and a motor is selected one standard size above that ($7.5\,\si{\kilo
\watt}$) to carry the duty with margin. The cubic-in-speed lesson of
worked example~5 recurs here as a square-in-flow lesson: the friction
head, and with it the power spent against friction, grows as $Q^2$, so
oversizing the duty flow is expensive.

\paragraph{Cavitation margin (NPSH).} The suction side is the
hazard. The net positive suction head available is the absolute
pressure head at the pump inlet above the vapour-pressure head:
\[
\text{NPSH}_a = \frac{p_{\text{atm}} - p_v}{\rho g} - z_s -
h_{f,\text{suction}},
\]
where $z_s = 2\,\si{\meter}$ is the suction lift and
$h_{f,\text{suction}}$ is the friction head in the suction run. With
$p_{\text{atm}} = 101.3\,\si{\kilo\pascal}$ and $p_v = 2.34\,\si{
\kilo\pascal}$,
\[
\frac{p_{\text{atm}} - p_v}{\rho g} = \frac{(101.3 - 2.34)\times10^3}{
998\times9.81} \approx 10.1\,\si{\meter}.
\]
The suction-run friction head is the major plus minor loss over the
$6\,\si{\meter}$ suction pipe with its entrance and one elbow
($\sum K_s \approx 0.5 + 0.3 = 0.8$):
\[
h_{f,\text{suction}} = \left(f\frac{L_s}{D} + \sum K_s\right)\frac{
\bar u^2}{2g} = \left(0.0176\frac{6}{0.150} + 0.8\right)\times0.101
\approx (0.70 + 0.8)\times0.101 \approx 0.15\,\si{\meter}.
\]
So
\[
\text{NPSH}_a = 10.1 - 2.0 - 0.15 \approx 7.9\,\si{\meter}.
\]
The pump's required suction head $\text{NPSH}_r$ is read from the
manufacturer's curve at the duty flow; for a pump of this size and
speed it is typically $3$ to $4\,\si{\meter}$. Take $\text{NPSH}_r =
3.5\,\si{\meter}$. The margin is
\[
\text{NPSH}_a - \text{NPSH}_r \approx 7.9 - 3.5 = 4.4\,\si{\meter},
\]
comfortably positive. The standard hydraulic-design practice is a
margin of at least $0.5$ to $1\,\si{\meter}$, or a ratio
$\text{NPSH}_a/\text{NPSH}_r \gtrsim 1.3$, so this installation is
safe against cavitation \cite{text:ch02t3-karassik-pumps}. Were the
sump drawn down in service, or the suction line fouled, the margin
would shrink, which is why the suction side is kept short, large in
bore, and free of throttling fittings.

\paragraph{Verdict.} The transfer line carries $25\,\si{\liter\per
\second}$ of water at $\bar u = 1.41\,\si{\meter\per\second}$,
$\text{Re}_D = 2.1\times10^5$, with a friction factor $f = 0.0176$, a
total friction head of $2.0\,\si{\meter}$ on a static lift of
$18\,\si{\meter}$, met by a single-stage pump trimmed to a duty head
of $20\,\si{\meter}$, drawing $6.5\,\si{\kilo\watt}$ at the shaft
behind a $7.5\,\si{\kilo\watt}$ motor, with a cavitation margin of
$4.4\,\si{\meter}$. Every number in the verdict was carried from the
fluid property and the geometry through the chain of relations, and
each link is a relation from earlier in the chapter. A design that
skips the regime check picks the wrong friction law; a design that
skips the minor losses underestimates the head on a fitting-heavy
line; a design that skips the NPSH check installs a pump that
cavitates and erodes its impeller within a season. The end-to-end
discipline is to carry the chain unbroken.

\begin{mastery}
\textbf{The Falkner-Skan family: boundary layers with pressure
gradient.} The Blasius solution is the zero-pressure-gradient member
of a one-parameter family of similarity solutions discovered by
Falkner and Skan (1931). For an external flow whose edge velocity
varies as a power of distance, $U_e(x) = C x^m$, the boundary-layer
equations still admit a similarity reduction. With the similarity
variable $\eta = y\sqrt{(m+1)U_e/(2\nu x)}$ and stream function
$\psi = \sqrt{2\nu x\,U_e/(m+1)}\,f(\eta)$, the partial differential
equation collapses to the Falkner-Skan ordinary differential
equation
\[
f''' + f f'' + \beta\,[1 - (f')^2] = 0,
\]
with $\beta = 2m/(m+1)$ the dimensionless pressure-gradient parameter
and boundary conditions $f(0) = f'(0) = 0$, $f'(\infty) = 1$. The
edge velocity of an inviscid flow over a wedge of total opening angle
$\pi\beta$ has exactly this power-law form, so $\beta$ is the wedge
parameter; $\beta = 0$ ($m=0$) recovers the flat-plate Blasius case,
$\beta = 1$ is plane stagnation-point flow, and positive $\beta$ is a
favourable (accelerating) gradient that thins the layer and fills the
profile. Negative $\beta$ is an adverse (decelerating) gradient that
thins the wall shear; the wall shear $f''(0)$ falls with decreasing
$\beta$ and reaches zero at $\beta \approx -0.1988$, the point of
incipient separation. Below that value no attached similarity
solution exists, which is the analytical statement that a strong
enough adverse pressure gradient separates the boundary layer. The
single parameter $\beta$ thus carries the whole qualitative story of
section~12.5: favourable gradients stabilise and thin, adverse
gradients thicken and separate, and the separation threshold is a
definite number.
\end{mastery}

\section{End-to-end case study: sizing a hydrodynamic journal bearing}\pathtag{standard}

The pump case study ran the chapter's relations through an external,
high-Reynolds-number flow. The same discipline applies to the opposite
regime, a thin viscous film at low Reynolds number where viscosity
carries the entire load. We size the main journal bearing of a
mid-size industrial pump or turbine rotor, working the chain duty load
$\to$ geometry $\to$ Sommerfeld number $\to$ eccentricity $\to$
minimum film thickness $\to$ friction power $\to$ oil-temperature rise,
and carrying the numbers at every link. The geometry is the one drawn
in figure~\ref{fig:vol03ch12:journal-bearing}, and the design point is
read off figure~\ref{fig:vol03ch12:sommerfeld-curve}.

\paragraph{The duty and the geometry.} A horizontal rotor weighs
$8.0\,\text{kN}$ on each of two journals, so each bearing carries a
radial load $W = 8.0\,\text{kN}$. The shaft journal has diameter
$D = 2r = 100\,\si{\milli\meter}$ ($r = 50\,\si{\milli\meter}$) and the
bearing is $L = 75\,\si{\milli\meter}$ long, giving a projected area
$A = L D = 0.075\times0.100 = 7.5\times10^{-3}\,\si{\meter\squared}$.
The shaft turns at $N = 1500\,\text{rev/min} = 25\,\text{rev/s}$, an
angular speed $\omega = 2\pi N = 157\,\si{\radian\per\second}$, so the
surface speed is $U = \omega r = 157\times0.050 = 7.85\,\si{\meter\per
\second}$. The bearing is babbitt-lined and fed with an ISO VG 32
turbine oil whose viscosity at the operating film temperature
(estimated at $60\,\degC$) is $\mu = 1.9\times10^{-2}\,\si{\pascal
\second}$, with density $\rho \approx 860\,\si{\kilogram\per\meter
\cubed}$. The diametral clearance is set by the common rule of one
thousandth of the diameter, so the radial clearance is $c = 0.001\,r
\times\,$(factor); we take $c = 0.050\,\si{\milli\meter}$ (a
clearance ratio $r/c = 1000$, mid-range for this size).

\paragraph{Bearing pressure and the lubrication regime.} The unit
load on the projected area is
\[
P = \frac{W}{LD} = \frac{8.0\times10^{3}}{7.5\times10^{-3}} \approx
1.07\times10^{6}\,\si{\pascal} \approx 1.07\,\text{MPa},
\]
within the $1$ to $3\,\text{MPa}$ band that babbitt bearings carry
comfortably \cite{text:stachowiak-batchelor}. The film Reynolds number
sets the lubrication theory: with the film thickness $h \sim c =
5\times10^{-5}\,\si{\meter}$ and the surface speed $U = 7.85\,\si{
\meter\per\second}$,
\[
\text{Re}_{\text{film}} = \frac{\rho U h}{\mu} = \frac{860\times7.85
\times5\times10^{-5}}{1.9\times10^{-2}} \approx 18,
\]
well below the value at which inertia matters, so the thin-film
Reynolds equation (creeping-flow lubrication, inertia neglected)
governs the film. This is the same approximation family as the
Couette and Stokes flows of section~12.4: a viscous-dominated flow in
a gap thin compared to its extent.

\paragraph{Sommerfeld number and eccentricity.} The dimensionless
group that orders journal-bearing performance is the Sommerfeld
number,
\[
S = \frac{\mu N}{P}\left(\frac{r}{c}\right)^2 = \frac{1.9\times10^{-2}
\times25}{1.07\times10^{6}}\times(1000)^2 = \frac{0.475}{1.07\times
10^{6}}\times10^{6} \approx 0.44.
\]
A more conservative design that runs the bearing at a lower
Sommerfeld number, by a tighter clearance or a heavier load, sits
nearer the steep part of the curve. For a finite bearing of length
ratio $L/D = 0.75$, the short-bearing and tabulated solutions of the
Reynolds equation give an eccentricity ratio in the range
$\varepsilon \approx 0.6$ to $0.7$ at this Sommerfeld number
\cite{text:stachowiak-batchelor}; the design chart in
figure~\ref{fig:vol03ch12:sommerfeld-curve} reads $\varepsilon
\approx 0.62$ off the trend
curve at $S\approx0.44$ for the duty point. We take $\varepsilon =
0.62$.

\paragraph{Minimum film thickness.} The film converges from
$h_{\max} = c(1+\varepsilon)$ on the unloaded side to a minimum
\[
h_{\min} = c(1-\varepsilon) = 5\times10^{-5}\times(1-0.62) \approx
1.9\times10^{-5}\,\si{\meter} = 19\,\si{\micro\meter}.
\]
The design check is that this minimum film clears the combined surface
roughness of journal and bush. A ground journal at $R_a \approx
0.4\,\si{\micro\meter}$ and a bored babbitt bush at $R_a \approx
0.8\,\si{\micro\meter}$ have a combined roughness of order
$1\,\si{\micro\meter}$; the working rule keeps the film thickness at
least three to five times the combined roughness
\cite{text:stachowiak-batchelor}, so a $19\,\si{\micro\meter}$ film
against a $1\,\si{\micro\meter}$ roughness gives a ratio near $19$,
comfortably in full-film (hydrodynamic) lubrication. The bearing does
not touch metal to metal in steady running.

\paragraph{Friction power and oil flow.} The friction torque comes
from the shear in the film. For a moderately loaded bearing the
friction coefficient referred to the load is of order
$f_b \approx (r/c)^{-1}\,(2\pi^2 S + \text{eccentricity terms})$; the
tabulated Petroff-corrected value at $S\approx0.44$, $\varepsilon=
0.62$ is $f_b \approx 3.5\times10^{-3}$
\cite{text:stachowiak-batchelor}. The friction force is then
$F_f = f_b W = 3.5\times10^{-3}\times8.0\times10^{3} \approx 28\,\si{
\newton}$, and the friction power dissipated in the film is
\[
P_f = F_f U = 28\times7.85 \approx 2.2\times10^{2}\,\si{\watt}
\approx 0.22\,\text{kW}.
\]
A Petroff estimate gives the upper bound for a lightly loaded,
concentric bearing: $P_{f,\text{Petroff}} = \frac{2\pi^2\mu N L D}{c}
\,r^2\omega/\,$(grouping), which for these numbers lands within a
factor of two of the loaded estimate; the loaded bearing dissipates
less because the film thins on the loaded side and thickens on the
other. Take $P_f \approx 0.22\,\text{kW}$ per bearing.

\paragraph{Oil-temperature rise.} The friction power heats the oil,
and the oil flow through the bearing carries the heat away. If the
bearing is fed $Q_{\text{oil}} \approx 0.6\,\text{L/min} = 1.0\times
10^{-5}\,\si{\meter\cubed\per\second}$, the temperature rise across
the bearing is set by an energy balance $P_f = \rho\,c_p\,Q_{\text{oil}}
\,\Delta T$, with oil specific heat $c_p \approx 2.0\times10^{3}\,\si{
\joule\per\kilogram\per\kelvin}$:
\[
\Delta T = \frac{P_f}{\rho\,c_p\,Q_{\text{oil}}} = \frac{220}{860
\times2.0\times10^{3}\times1.0\times10^{-5}} \approx 13\,\si{\kelvin}.
\]
A $13\,\si{\kelvin}$ rise on a $47\,\degC$ inlet gives a $60\,\degC$
film temperature, consistent with the viscosity we assumed; a first
pass that guessed the film temperature is closed by checking that the
computed rise reproduces it, and one iteration on $\mu(T)$ converges.
The design rule keeps the outlet temperature below the oil's oxidation
limit (typically $70$ to $80\,\degC$ for mineral turbine oil), so the
flow rate is set to hold $\Delta T$ within that ceiling.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-journal-bearing}

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-sommerfeld-curve}

\paragraph{Verdict.} The journal carries $8.0\,\text{kN}$ at a unit
load of $1.07\,\text{MPa}$, running at a Sommerfeld number $S\approx
0.44$ and an eccentricity ratio $\varepsilon\approx0.62$, on a minimum
film of $19\,\si{\micro\meter}$ that clears the combined surface
roughness by a factor near $19$, dissipating $0.22\,\text{kW}$ of
friction carried away by $0.6\,\text{L/min}$ of oil at a $13\,\si{
\kelvin}$ rise. Every number descends from the load and the geometry
through the thin-film viscous relations, and each link is the
low-Reynolds-number companion of a relation from earlier in the
chapter: the Reynolds equation is the lubrication limit of
Navier-Stokes, the Sommerfeld number is the bearing's Reynolds-like
ordering parameter, and the film shear is Couette flow in a converging
gap. A design that skips the film-temperature iteration uses the wrong
viscosity and misreads the Sommerfeld number; a design that skips the
minimum-film check sets a clearance that drops the bearing into
boundary lubrication, where metal touches metal and the babbitt wipes
within hours of start-up. The end-to-end discipline is the same as the
pump's: carry the chain unbroken, and close every assumed quantity
against the result it produces.

\section{End-to-end case study: vehicle drag, power, and range}\pathtag{standard}

The pump and the bearing ran the chapter's relations through internal
flows. The third study runs them through the external flow over a
moving body, where the drag coefficient of section~12.5 meets the
rolling resistance of the tyres to set the power a vehicle spends, the
energy it spends per distance, and the range a given store of energy
buys. We work the chain frontal area $\to$ drag area $\to$ road-load
power $\to$ energy per distance $\to$ range, then run a drag-reduction
change through the same chain and read off what it is worth, carrying
the numbers at every link \cite{text:white-fluid-mechanics,
text:anderson-aerodynamics}.

\paragraph{The vehicle and the air.} A mid-size electric car has a
mass $m = 1500\,\si{\kilogram}$, a frontal area $A = 2.2\,\si{\meter
\squared}$, and a drag coefficient $C_D = 0.30$, giving a drag area
\[
C_D A = 0.30\times2.2 = 0.66\,\si{\meter\squared}.
\]
The drag area, not the bare coefficient, is the quantity that sets the
force; a sleek coefficient on a large frontal area can lose to a
boxy coefficient on a small one. The air is at sea level,
$\rho = 1.2\,\si{\kilogram\per\meter\cubed}$. The tyres roll with a
coefficient of rolling resistance $C_{rr} = 0.012$, typical for a
modern passenger tyre on smooth tarmac. The battery stores
$E_{\text{batt}} = 60\,\si{\kilo\watt\hour} = 2.16\times10^{8}\,\si{
\joule}$, and the drivetrain delivers it to the road at an efficiency
$\eta = 0.88$ from battery to wheel.

\paragraph{The two road loads.} Two forces oppose steady cruise on the
level. The aerodynamic drag grows as the square of speed,
\[
F_{\text{aero}} = \tfrac12\rho\,C_D A\,U^2,
\]
and the rolling resistance is nearly constant with speed,
\[
F_{\text{roll}} = C_{rr}\,m g = 0.012\times1500\times9.81 \approx
177\,\si{\newton}.
\]
At a cruise of $U = 27.8\,\si{\meter\per\second}$ ($100\,\text{km/h}$)
the aerodynamic force is
\[
F_{\text{aero}} = 0.5\times1.2\times0.66\times27.8^2 \approx 306\,\si{
\newton},
\]
so the two road loads are comparable at this speed, with aerodynamics
already the larger. The total road-load force is $F = F_{\text{aero}}
+ F_{\text{roll}} \approx 306 + 177 = 483\,\si{\newton}$.

\paragraph{Road-load power and the crossover speed.} The power at the
wheels is the force times the speed, so the rolling part grows
linearly and the aerodynamic part grows as the cube:
\[
P(U) = C_{rr}m g\,U + \tfrac12\rho\,C_D A\,U^3.
\]
At $100\,\text{km/h}$,
\[
P = 177\times27.8 + 0.5\times1.2\times0.66\times27.8^3 \approx 4.9
\times10^{3} + 8.5\times10^{3} \approx 1.34\times10^{4}\,\si{\watt}
\approx 13.4\,\si{\kilo\watt}.
\]
The two terms cross where $C_{rr}m g\,U = \tfrac12\rho C_D A\,U^3$,
that is at
\[
U_{\text{cross}} = \sqrt{\frac{2 C_{rr}m g}{\rho\,C_D A}} = \sqrt{
\frac{2\times177}{1.2\times0.66}} \approx \sqrt{447} \approx 21.1\,
\si{\meter\per\second},
\]
about $76\,\text{km/h}$. Below the crossover the tyres dominate the
loss; above it the air dominates, and the cubic term pulls away
steeply. The split is drawn in figure~\ref{fig:vol03ch12:road-load}.
This crossover is why a drag change is felt on the motorway and barely
felt in town, and why hybrid city driving rewards low rolling
resistance while motorway range rewards a low drag area.

\input{volumes/03-force/ch12-viscous-flow-boundary-layers/figures/fig-road-load}

\paragraph{Energy per distance and range.} The energy drawn from the
battery per unit distance is the wheel force divided by the drivetrain
efficiency,
\[
\frac{E}{d} = \frac{F}{\eta} = \frac{C_{rr}m g + \tfrac12\rho C_D A\,
U^2}{\eta}.
\]
At $100\,\text{km/h}$,
\[
\frac{E}{d} = \frac{483}{0.88} \approx 549\,\si{\joule\per\meter} =
549\,\si{\kilo\joule\per\kilo\meter} \approx 0.153\,\si{\kilo\watt
\hour\per\kilo\meter},
\]
about $15.3\,\si{\kilo\watt\hour}$ per $100\,\si{\kilo\meter}$, in
line with the published consumption of cars of this class. The range
on the full battery is
\[
R = \frac{E_{\text{batt}}}{E/d} = \frac{2.16\times10^{8}}{549}
\approx 3.9\times10^{5}\,\si{\meter} \approx 393\,\si{\kilo\meter}.
\]

\paragraph{The drag-reduction change, carried through.} A redesign
trims the drag coefficient from $0.30$ to $0.26$, a $13\,\%$
reduction, lowering the drag area to $C_D A = 0.26\times2.2 = 0.572\,
\si{\meter\squared}$. Run it through the same chain at the same
$100\,\text{km/h}$. The aerodynamic force falls to
\[
F_{\text{aero}}' = 0.5\times1.2\times0.572\times27.8^2 \approx 265\,
\si{\newton},
\]
a drop of $41\,\si{\newton}$, while the rolling force is unchanged.
The total road load falls to $F' = 265 + 177 = 442\,\si{\newton}$, a
$8.5\,\%$ reduction in the force and so in the energy per distance,
which becomes $E'/d = 442/0.88 \approx 502\,\si{\joule\per\meter}$.
The range rises to
\[
R' = \frac{2.16\times10^{8}}{502} \approx 4.30\times10^{5}\,\si{\meter}
\approx 430\,\si{\kilo\meter},
\]
a gain of $37\,\si{\kilo\meter}$, or $9.4\,\%$, from a drag-coefficient
change of $13\,\%$. The range gain is smaller than the
drag-coefficient cut because rolling resistance carries part of the
load that the redesign does not touch; the aerodynamic share of the
saving is diluted by the constant rolling term. At $120\,\text{km/h}$,
where the cubic aerodynamic term is a larger fraction of the total,
the same drag cut would return a larger relative range gain, which is
the quantitative form of the lesson that drag matters most at speed.

\paragraph{Verdict.} The car spends $13.4\,\si{\kilo\watt}$ at the
wheels at $100\,\text{km/h}$, split $8.5\,\si{\kilo\watt}$ to the air
and $4.9\,\si{\kilo\watt}$ to the tyres, draws $0.153\,\si{\kilo\watt
\hour\per\kilo\meter}$ from the battery, and runs $393\,\si{\kilo
\meter}$ on a $60\,\si{\kilo\watt\hour}$ charge; a drag-coefficient
cut from $0.30$ to $0.26$ buys $37\,\si{\kilo\meter}$ of range. Every
number descends from the drag area and the rolling coefficient
through the road-load chain, and each link is a relation from
section~12.5: the drag coefficient sets the aerodynamic force, the
force times speed is the power, the power integrated over distance is
the energy, and the stored energy divided by the energy per distance
is the range. A design that quotes the drag coefficient without the
frontal area misstates the force; a design that counts aerodynamics
without rolling resistance overstates the range gain from a body
change; a design that ignores the cubic speed dependence sizes the
motor for the wrong cruise. The end-to-end discipline is the pump's
and the bearing's once more: carry the chain unbroken from the
dimensionless coefficient to the number a driver reads off the range
display.

\section{Failure: separation, stall, and unexpected drag}\pathtag{core}

Every chapter in this work has a failure section. The viscous-flow
failure modes cluster into three families: boundary-layer
separation under adverse pressure gradient, transition-related
phenomena that change the drag coefficient discontinuously across
a Reynolds-number range, and turbulence-induced fluctuating loads
that drive resonant structural response.

\subsection{Boundary-layer separation and airfoil stall}

An airfoil at increasing angle of attack develops an increasing
adverse pressure gradient over the rear of its upper surface. The
boundary layer thickens, slows, and eventually reverses; the flow
separates from the surface and forms a large, low-pressure wake.
The lift, which had been growing linearly with angle of attack,
peaks and then drops sharply at the stall angle. The post-stall
drag is several times the pre-stall drag.

The stall angle for a typical conventional airfoil is $12^{\circ}$
to $20^{\circ}$, depending on Reynolds number, leading-edge
geometry, and surface roughness. High-lift devices on aircraft
wings (slats, slotted flaps, vortex generators) delay separation
and extend the usable angle-of-attack range; the working principle
is to inject high-energy flow into the boundary layer to overcome
the adverse pressure gradient near the trailing edge.

Stall is the canonical aerodynamic failure mode. Air France Flight
447, on 1 June 2009, entered a deep stall at $35{,}000\,\text{ft}$
after the autopilot disengaged on inconsistent airspeed data; the
crew did not recognise the stalled state, and the aircraft
descended in a stall until impact, with the loss of all $228$
people aboard \cite{acc:bea-af447}. The stall at altitude is
particularly hazardous: the recovery requires a substantial loss
of altitude to regain airspeed, the aerodynamic control
authority is degraded in the wake of the stalled wing, and the
cockpit instrumentation under unreliable airspeed conditions does
not give the crew an unambiguous indicator of the stalled state.
The AF447 lesson for boundary-layer analysis is that stall is a
flight-envelope boundary, not a post-hoc engineering observation.
The working aircraft certification discipline embeds stall-warning
systems, stick shakers, stall recovery procedures, and crew
training in unreliable-airspeed scenarios. The post-AF447
regulatory response included revised crew-training requirements
for unreliable airspeed and high-altitude upset, and changes to
autoflight handoff design \cite{acc:bea-af447}.

\subsection{Diffuser stall}

A diverging duct (diffuser) that tries to decelerate a flow and
recover its dynamic pressure can suffer separation when the
divergence angle is too large or the inlet velocity profile is too
non-uniform. The separated flow leaves the diffuser walls and
forms a free jet through the centre of the diffuser, with
recirculating regions in the corners. The pressure recovery drops
sharply; the diffuser loses its design function.

The threshold divergence angle is approximately $7^{\circ}$ on
each side for unaided flow (a $14^{\circ}$ total included angle).
Above this, separation is likely; below it, the boundary layer
typically stays attached. The discipline of diffuser design uses
this empirical limit, supplemented by computational and
experimental verification for specific geometries.

The trade-off the angle limit encodes is between length and recovery.
A diffuser that recovers a given area ratio with a gentle angle is
long and heavy and accumulates wall-friction loss over its length; a
short diffuser at a steep angle separates and loses far more in the
separated wake than it saves in friction. The optimum sits just below
the separation boundary, which is why the $7^{\circ}$ rule is so
durable: it is the steepest angle that keeps the flow attached for an
ordinary inlet profile, and therefore the shortest diffuser that
recovers its design pressure. When the application cannot afford the
length, the design buys attachment back by energising the boundary
layer, with vortex generators that mix high-momentum core flow into
the near-wall region, with boundary-layer suction through slots in the
wall, or with splitter vanes that subdivide the duct into several
narrow-angle passages. Each of these is the diffuser version of the
high-lift devices that delay airfoil stall, and each works by the same
mechanism: feed momentum into the near-wall fluid so it survives the
adverse pressure gradient to the diffuser exit.

\subsection{Vortex-induced vibration}

A bluff body (a cylinder, a chimney, a suspension-bridge cable) in
cross flow sheds Kármán vortices at the Strouhal frequency. The
shedding produces an oscillating lift on the body, which excites
resonant vibration when the shedding frequency approaches a
natural frequency of the body. The amplitude can grow to levels
that produce fatigue failure or, in extreme cases, structural
collapse.

The original Tacoma Narrows suspension bridge collapse of
$7$ November $1940$ \cite{acc:tacoma-narrows-1941} is the
canonical large-scale case (also treated in chapter 6 of this
volume on vibrations and chapter 13 on bluff-body aerodynamics).
The deck's torsional self-excitation under steady wind was not
classical vortex shedding alone but a related aeroelastic
phenomenon (torsional flutter); the chapter's narrow point is
that the viscous-flow physics of vortex shedding and the
aeroelastic-flutter physics of bridge decks are both governed by
the same Reynolds-number-dependent separated-flow regimes.

Industrial cases of vortex-induced vibration: power-line galloping
in ice-coated conductors (where the asymmetric ice cross-section
produces an instability beyond simple Kármán shedding); marine
riser vibration in subsea oil and gas production; resonance of
exhaust stacks of power stations under moderate cross winds; and
acoustic resonance of vehicle door seals at highway speeds. The
working design discipline is to compute the Strouhal frequency at
the operating flow speed, check it against the structure's natural
frequencies, and either shift the structural frequency or modify
the cross-section to disrupt the vortex shedding.

\subsection{The drag crisis surprise}

A bluff body whose operating Reynolds number sits near the
drag-crisis transition is at risk of dramatic drag changes from
small perturbations. A circular cylinder at $\text{Re} = 2\times
10^5$ has $C_D \approx 0.5$; the same cylinder at $\text{Re} =
4\times 10^5$ has $C_D \approx 0.2$. A design that sits near the
transition can experience a factor-of-two change in drag from
small changes in surface roughness, freestream turbulence, or
operating speed.

The discipline is to know whether the design Reynolds number is
subcritical, supercritical, or near transition, and to design for
the bracket of possible $C_D$ values rather than for a single
point estimate. A wind-tunnel test in a low-turbulence facility
may under-predict the full-scale $C_D$ if the model sits in
subcritical conditions while the full-scale operates supercritical.

\subsection{The diagnostic principle for viscous-flow failures}

The viscous-flow diagnostic question is, at every body in any
flow:

\begin{itemize}[leftmargin=*]
  \item What is the Reynolds number? Laminar, transitional, or
    turbulent regime?
  \item What is the local pressure gradient on the body? Favourable
    (boundary layer thins), zero (Blasius), or adverse (separation
    risk)?
  \item Where is the separation point likely to be? Has it been
    measured or predicted by a validated tool?
  \item What is the post-separation wake structure? Steady, with
    vortex shedding, with reattachment?
  \item Is the Strouhal frequency at the design speed coincident
    with a structural natural frequency?
  \item Is the body near a drag-crisis transition?
\end{itemize}

A design that has answered all six is a viscous-flow-aware design.
A design that has answered none of the six is in the regime where
real-flow drag and lift can depart from the inviscid prediction by
factors of two or more, and the difference shows up as an
operating-cost surprise, a structural-fatigue surprise, or, in the
worst case, a stability or control surprise of the kind that
brings airplanes down.

\begin{principle}[Real-flow drag has two parts; both are Reynolds-number-dependent]
The drag on a body in viscous flow is the sum of skin-friction
drag (from the integrated viscous shear stress) and pressure drag
(from the integrated surface pressure). The relative importance
depends on the body's shape: streamlined bodies are
friction-dominated, bluff bodies are pressure-dominated. Both
components vary with Reynolds number, sometimes discontinuously
(the drag crisis). The discipline is to compute the Reynolds number
first, identify the regime, and use the regime-appropriate drag
model rather than an inviscid extrapolation.
\end{principle}

\begin{archetype}[Transport, scaling, failure: the boundary-layer triad]
The viscous-flow archetype puts three relations to work. The
Reynolds number determines the regime (laminar, transitional,
turbulent); the boundary-layer thickness scales as $\sqrt{\nu x/
U_\infty}$ (laminar) or $x \cdot \text{Re}_x^{-1/5}$ (turbulent);
the wall shear stress sets the skin-friction contribution to drag.
The archetype's failure mode is separation, triggered by adverse
pressure gradient; the failure's consequences are stall, drag
spikes, vortex shedding, and aeroelastic instability. Every
viscous-flow analysis works through these three: regime,
boundary-layer scaling, separation risk.
\end{archetype}

\begin{estimation}
\textbf{Estimate the skin-friction drag on a Boeing 737 fuselage at
cruise.}

\smallskip
\textit{Estimate, before reading on.}

\smallskip
A working estimate goes through the following moves. A Boeing
$737$ fuselage has a length of about $38\,\si{\meter}$ and a
diameter of about $3.8\,\si{\meter}$. Treat the fuselage as a
cylinder for the wetted-area estimate: the curved surface area is
approximately $\pi D L \approx 3.14 \times 3.8 \times 38
\approx 450\,\si{\meter}^2$. The full wetted surface including
nose, tail, wings, and tail surfaces is roughly twice the
fuselage value, so total wetted area $S_w \approx 900\,\si{\meter}
^2$.

At cruise (Mach $0.78$, altitude $35{,}000\,\text{ft}$), the
freestream is $U_\infty \approx 230\,\si{\meter\per\second}$,
the air density $\rho \approx 0.38\,\si{\kilogram\per\meter\cubed}$,
the air kinematic viscosity $\nu \approx 4\times 10^{-5}\,\si{\meter
}^2/\si{\second}$. The length-based Reynolds number is
\[
\text{Re}_L = \frac{U_\infty L}{\nu} = \frac{230 \times 38}{4
\times 10^{-5}} \approx 2.2\times 10^8.
\]
The boundary layer is fully turbulent over nearly the entire
fuselage; the average skin-friction coefficient is
\[
\bar{C}_f \approx \frac{0.074}{\text{Re}_L^{1/5}} \approx
\frac{0.074}{(2.2\times 10^8)^{0.2}} \approx \frac{0.074}{47}
\approx 0.0016.
\]
The skin-friction drag is then
\[
F_D = \tfrac{1}{2}\rho U_\infty^2 \cdot S_w \cdot \bar{C}_f
\approx 0.5 \times 0.38 \times 230^2 \times 900 \times 0.0016
\approx 14{,}500\,\si{\newton}.
\]
A working figure for total $737$ cruise drag is about $35\,\text{kN}$
to $45\,\text{kN}$; skin friction accounts for roughly half, with
the rest divided between induced (vortex-related) drag, pressure
drag at the rear, and miscellaneous excrescence drag.

The postmortem: the skin-friction estimate is within a factor of
two of the working figure, despite all the simplifications
(uniform $C_f$, cylinder approximation, neglect of wing thickness).
The factor-of-two-or-better accuracy is characteristic of
boundary-layer estimates done with the integrated turbulent
correlations; the same accuracy is what makes the relations useful
for first-pass design. A high-accuracy prediction requires
either wind-tunnel measurement or validated CFD with appropriate
turbulence model.
\end{estimation}

\section{Project}\pathtag{core}

\begin{project}[Measure pressure drop across a pipe; verify against the Moody chart]
The reader instruments a household pipe section, measures the
pressure drop as a function of flow rate, computes the friction
factor, and compares to the Moody chart for the pipe's
roughness. The deliverable is a twenty- to thirty-page engineering
report.

\textbf{Track}. Physical build plus measurement plus analysis.
\textbf{Hazard class}: low.

\textbf{Apparatus}. A length of household pipe (smooth copper,
PVC, or PEX) of known diameter ($15$ to $25\,\si{\milli\meter}$
typical), length of $1$ to $2\,\si{\meter}$, with pressure taps at
the inlet and outlet of the measured section. Pressure measurement
by U-tube manometer (water, oil, or mercury depending on the
pressure range), differential pressure transducer, or two
inexpensive digital pressure gauges. Flow rate measurement by
collecting the outflow over a measured time interval.

\textbf{Procedure}. The reader varies the flow rate over the
widest available range (from a low trickle to the maximum supply),
records the pressure drop and the volume flow rate at each
setting, and computes:

\begin{itemize}[leftmargin=*]
  \item The mean flow velocity $\bar{u} = Q/(\pi D^2/4)$.
  \item The Reynolds number $\text{Re}_D = \bar{u} D/\nu$, using
    water's kinematic viscosity at the measured temperature.
  \item The friction factor $f = (\Delta p \cdot D)/(L \cdot
    \tfrac{1}{2}\rho \bar{u}^2)$.
  \item A log-log plot of $f$ vs.\ $\text{Re}_D$, overlaid on the
    laminar curve $f = 64/\text{Re}_D$ and on the Colebrook-White
    or Haaland approximation for the pipe's estimated roughness.
\end{itemize}

\textbf{Analysis}. The reader identifies:

\begin{itemize}[leftmargin=*]
  \item The Reynolds-number range in which the data fall (laminar,
    transitional, turbulent).
  \item The agreement (or disagreement) with the Moody-chart
    prediction at each $\text{Re}_D$.
  \item The estimated pipe roughness $\epsilon$ that best fits the
    data, compared to handbook values for the pipe material.
  \item Sources of error: pressure-tap geometry, entrance and exit
    effects (the data should be in the fully developed region,
    not in the developing flow near the pipe inlet), bend effects
    if any, flow-rate measurement uncertainty.
\end{itemize}

\textbf{Reflection ($1{,}500$ to $3{,}000$ words)}. The reader
addresses:

\begin{itemize}[leftmargin=*]
  \item The accuracy with which the Moody chart predicts the
    measured friction factor across the operating range.
  \item The Reynolds-number boundary at which the data depart from
    the laminar prediction, compared to the textbook value of
    $\text{Re}_D = 2300$.
  \item One source of measurement uncertainty whose impact the
    reader was able to bound (or could not), with a brief
    discussion of why.
\end{itemize}

\textbf{Deliverable}. The pipe-instrument setup (photographed and
dimensioned), the raw data, the analysis, and the reflection.

\textbf{Time}. Two to four hours of apparatus construction, three
to six hours of measurement, four to eight hours of analysis and
write-up.
\end{project}

\section{Exercises}

The exercises mix calculation, derivation, estimation, simulation,
design, diagnosis, failure analysis, and judgment.

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
Water at $20\,\degC$ ($\rho = 998\,\si{\kilogram\per\meter\cubed}$,
$\mu = 1.0\times 10^{-3}\,\si{\pascal\second}$) flows through a
$25\,\si{\milli\meter}$ diameter pipe at $0.5\,\si{\meter\per
\second}$. Find the Reynolds number and identify the flow regime.

\paragraph{Solution sketch.} $\text{Re}_D = \rho\bar u D/\mu =
998\times0.5\times0.025/(1.0\times10^{-3}) \approx 1.25\times10^{4}$.
This is well above $4000$, so the flow is turbulent. (The laminar
ceiling $\text{Re}_D\approx2300$ would correspond here to
$\bar u\approx0.09\,\si{\meter\per\second}$, so any ordinary tap
flow in this pipe is turbulent.)
\end{exercise}

\begin{exercise}
For the same pipe and fluid as above, find the pressure drop per
metre of length in the fully developed laminar regime. Then
compute the pressure drop in turbulent flow at $\bar{u} = 3\,\si{
\meter\per\second}$ for a smooth pipe ($\epsilon \approx 0$), using
a friction factor from the explicit Haaland correlation
$f^{-1/2} = -1.8\log_{10}[(\epsilon/(3.7D))^{1.11} + 6.9/\text{Re}_D]$.

\paragraph{Solution sketch.} Laminar at $\bar u = 0.5\,\si{\meter
\per\second}$: $\text{Re}_D \approx 1.25\times10^4$ is actually
turbulent, so the laminar formula does not apply at this speed. Take
the laminar case at a speed inside the laminar range; the cleaner
reading of the prompt is to compute the Hagen-Poiseuille drop that
the laminar law \emph{would} give and the turbulent drop at
$3\,\si{\meter\per\second}$, then compare. Hagen-Poiseuille:
$\Delta p/L = 128\mu Q/(\pi D^4)$ with $Q = \bar u\,\pi D^2/4 =
0.5\times\pi\times0.025^2/4 = 2.45\times10^{-4}\,\text{m}^3/\text{s}$,
giving $\Delta p/L = 128\times10^{-3}\times2.45\times10^{-4}/(\pi
\times0.025^4) \approx 2.55\,\si{\pascal\per\meter}$ (equivalently
$f = 64/\text{Re}_D \approx 5.1\times10^{-3}$ with $\Delta p/L =
f\rho\bar u^2/(2D)$). Turbulent at $\bar u = 3\,\si{\meter\per
\second}$: $\text{Re}_D = 998\times3\times0.025/10^{-3} \approx
7.5\times10^{4}$; with $\epsilon\approx0$, Haaland gives
$f^{-1/2} = -1.8\log_{10}(6.9/7.5\times10^{4}) \approx 7.40$, so
$f \approx 0.0183$ and $\Delta p/L = f\rho\bar u^2/(2D) =
0.0183\times998\times9/(2\times0.025) \approx 3.3\times10^{3}\,\si{
\pascal\per\meter}$. The turbulent drop per metre is roughly three
orders of magnitude larger, partly because $\Delta p\propto\bar u^2$
turbulent versus $\bar u$ laminar, and partly the $36\times$ speed-
squared factor.
\end{exercise}

\begin{exercise}
A flat plate of length $1\,\si{\meter}$ is in a freestream of air
($\rho = 1.225\,\si{\kilogram\per\meter\cubed}$, $\nu = 1.5\times
10^{-5}\,\si{\meter}^2/\si{\second}$) at $20\,\si{\meter\per
\second}$. Find $\text{Re}_L$, the Blasius boundary-layer thickness
$\delta$ at the trailing edge, and the average skin-friction
coefficient.

\paragraph{Solution sketch.} $\text{Re}_L = U_\infty L/\nu =
20\times1/(1.5\times10^{-5}) = 1.33\times10^{6}$. This exceeds the
smooth-plate transition near $5\times10^5$, so most of the plate is
turbulent; the laminar Blasius formulas are an approximation valid
only over the front. Using the laminar relations as the prompt asks:
$\delta = 5.0\,L/\sqrt{\text{Re}_L} = 5.0/\sqrt{1.33\times10^6}
\approx 4.3\times10^{-3}\,\si{\meter}$ and $\bar C_f = 1.328/\sqrt{
\text{Re}_L} = 1.328/1153 \approx 1.15\times10^{-3}$. (For the
turbulent estimate one would use $\delta = 0.37\,L\,\text{Re}_L^{-1/5}
\approx 0.022\,\si{\meter}$ and $\bar C_f = 0.074\,\text{Re}_L^{-1/5}
\approx 4.4\times10^{-3}$, several times the laminar values; the
honest answer notes the plate is past transition.)
\end{exercise}

\begin{exercise}
A sphere of $5\,\si{\milli\meter}$ diameter falls through glycerin
($\rho = 1260\,\si{\kilogram\per\meter\cubed}$, $\mu = 1.5\,\si{
\pascal\second}$) under gravity. Find the terminal velocity using
the Stokes drag law. Check that the resulting Reynolds number is
in the Stokes regime ($\text{Re} < 1$).

\paragraph{Solution sketch.} Need the sphere density; take a steel
sphere $\rho_s \approx 7800\,\si{\kilogram\per\meter\cubed}$.
Balance weight minus buoyancy against Stokes drag:
$U = (\rho_s - \rho_f)g\,d^2/(18\mu)$ with $d = 5\,\si{\milli\meter}$.
$U = (7800-1260)\times9.81\times(5\times10^{-3})^2/(18\times1.5)
= 6540\times9.81\times2.5\times10^{-5}/27 \approx 5.9\times10^{-2}
\,\si{\meter\per\second}$. Check: $\text{Re} = \rho_f U d/\mu =
1260\times0.059\times0.005/1.5 \approx 0.25 < 1$, so the Stokes
regime holds and the answer stands. (Glycerin's high viscosity is
what keeps such a fast-settling steel ball in the creeping regime.)
\end{exercise}

\begin{exercise}
Air ($\nu = 1.5\times 10^{-5}\,\si{\meter}^2/\si{\second}$) flows at
$10\,\si{\meter\per\second}$ past a circular cylinder of
$50\,\si{\milli\meter}$ diameter. Compute the Reynolds number, the
Strouhal number ($\approx 0.21$ in this range), and the vortex-
shedding frequency.

\paragraph{Solution sketch.} $\text{Re} = UD/\nu = 10\times0.05/
(1.5\times10^{-5}) \approx 3.3\times10^{4}$, in the range where the
Strouhal number is roughly constant at $\text{St}\approx0.21$. The
shedding frequency is $f_s = \text{St}\,U/D = 0.21\times10/0.05 =
42\,\si{\hertz}$. (At $\text{Re}\approx3\times10^4$ the wake is a
fully turbulent vortex street; the $42\,\si{\hertz}$ tone is audible
as the aeolian whistle of a wire in wind.)
\end{exercise}

\begin{exercise}
A diffuser expands a $50\,\si{\milli\meter}$ diameter pipe to
$100\,\si{\milli\meter}$ diameter over a length of $300\,\si{\milli
\meter}$. Find the divergence half-angle and state whether the
diffuser is likely to suffer separation.

\paragraph{Solution sketch.} The radius grows from $25\,\si{\milli
\meter}$ to $50\,\si{\milli\meter}$ over $300\,\si{\milli\meter}$,
so the half-angle is $\theta = \arctan[(50-25)/300] = \arctan(0.0833)
\approx 4.8^{\circ}$. This is below the rule-of-thumb threshold of
about $7^{\circ}$ per side, so the boundary layer is expected to stay
attached and the diffuser should recover pressure reasonably well.
The margin is modest, so a non-uniform or already-thickened inlet
profile could still trigger local separation.
\end{exercise}

\begin{exercise}
A racing-car tyre rotates at $200\,\text{km/h}$ on
the road; the boundary layer on its outer surface is turbulent.
Estimate the boundary-layer thickness at a streamwise distance of
$0.3\,\si{\meter}$ on the tyre's leading face (assume air, sea-
level conditions).

\paragraph{Solution sketch.} Surface speed $U = 200\,\text{km/h} =
55.6\,\si{\meter\per\second}$; air $\nu = 1.5\times10^{-5}\,\si{
\meter}^2/\si{\second}$. Local Reynolds number at $x = 0.3\,\si{
\meter}$: $\text{Re}_x = Ux/\nu = 55.6\times0.3/(1.5\times10^{-5})
\approx 1.1\times10^{6}$, turbulent (consistent with the prompt).
Turbulent thickness $\delta = 0.37\,x\,\text{Re}_x^{-1/5} = 0.37
\times0.3\times(1.1\times10^6)^{-0.2} \approx 0.111\times0.0615
\approx 6.8\times10^{-3}\,\si{\meter}$, about $7\,\si{\milli\meter}$.
\end{exercise}

\begin{exercise}
A car of frontal area $2.2\,\si{\meter}^2$ travels at $30\,\si{
\meter\per\second}$ in still air. Its drag coefficient is $0.32$.
Find the drag force and the power dissipated against drag.

\paragraph{Solution sketch.} With $\rho = 1.2\,\si{\kilogram\per
\meter\cubed}$, $F_D = \tfrac12\rho U^2 A\,C_D = 0.5\times1.2\times
30^2\times2.2\times0.32 \approx 3.8\times10^{2}\,\si{\newton}$. Power
$P = F_D U = 380\times30 \approx 1.1\times10^{4}\,\si{\watt} =
11.4\,\text{kW}$. The cube-of-speed scaling means a $10\%$ speed
rise raises this power by about $33\%$.
\end{exercise}

\begin{exercise}
A long-span suspension-bridge cable of diameter $0.5\,\si{\meter}$
is in a $15\,\si{\meter\per\second}$ wind. Compute the Reynolds
number, the vortex-shedding frequency, and compare to a typical
cable natural frequency (state your assumption about the cable's
fundamental mode).

\paragraph{Solution sketch.} Air $\nu = 1.5\times10^{-5}\,\si{\meter
}^2/\si{\second}$. $\text{Re} = UD/\nu = 15\times0.5/(1.5\times
10^{-5}) = 5\times10^{5}$, in the range where $\text{St}\approx0.2$.
Shedding frequency $f_s = \text{St}\,U/D = 0.2\times15/0.5 = 6\,\si{
\hertz}$. A long suspension-bridge main cable has a low fundamental,
often a fraction of a hertz to a few hertz; assume $f_n\approx
0.5\,\si{\hertz}$ for a $100\,\si{\meter}$ free length. Then $f_s =
6\,\si{\hertz}\gg f_n$, so this wind speed excites a higher cable
mode rather than the fundamental; lock-in would occur at a lower
wind speed for the fundamental ($U_{\text{lock}} = f_n D/\text{St} =
0.5\times0.5/0.2 = 1.25\,\si{\meter\per\second}$) and at successively
higher speeds for higher modes.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the Hagen-Poiseuille velocity profile $u_z(r) = G(R^2 -
r^2)/(4\mu)$ from the steady cylindrical-coordinates Navier-Stokes
equation, stating the assumptions used.

\paragraph{Solution sketch.} Assume steady, incompressible,
axisymmetric, fully developed flow: $u_r = u_\theta = 0$,
$\partial u_z/\partial z = 0$, no swirl, constant properties, and a
constant axial pressure gradient $dp/dz = -G$. The $z$-momentum
Navier-Stokes equation in cylindrical coordinates collapses to
$\mu\,\tfrac1r\,\tfrac{d}{dr}(r\,du_z/dr) = -G$. Integrate once:
$r\,du_z/dr = -Gr^2/(2\mu) + C_1$. Symmetry at the axis,
$du_z/dr|_{r=0} = 0$, forces $C_1 = 0$. Integrate again:
$u_z = -Gr^2/(4\mu) + C_2$. No-slip at the wall, $u_z(R) = 0$,
gives $C_2 = GR^2/(4\mu)$, so $u_z(r) = G(R^2 - r^2)/(4\mu)$, the
parabola claimed.
\end{exercise}

\begin{exercise}
Show that the average velocity in Hagen-Poiseuille flow is half
the centreline velocity, and that the laminar friction factor is
$f = 64/\text{Re}_D$.

\paragraph{Solution sketch.} Average over the area:
$\bar u = \tfrac{1}{\pi R^2}\int_0^R u_z(r)\,2\pi r\,dr =
\tfrac{2}{R^2}\int_0^R \tfrac{G}{4\mu}(R^2 - r^2)\,r\,dr =
\tfrac{G}{2\mu R^2}[R^2 r^2/2 - r^4/4]_0^R = \tfrac{G}{2\mu R^2}
\cdot R^4/4 = GR^2/(8\mu)$. Since $u_{\max} = GR^2/(4\mu)$, indeed
$\bar u = \tfrac12 u_{\max}$. For the friction factor, write the
Hagen-Poiseuille drop as $\Delta p = 32\mu L\bar u/D^2$ (from
$G = 8\mu\bar u/R^2 = 32\mu\bar u/D^2$). Set this equal to the
Darcy-Weisbach form $\Delta p = f\tfrac{L}{D}\tfrac{\rho\bar u^2}{2}$
and solve: $f = 64\mu/(\rho\bar u D) = 64/\text{Re}_D$.
\end{exercise}

\begin{exercise}
Derive the Blasius scaling $\delta(x) \sim \sqrt{\nu x/U_\infty}$
by setting the inertial and viscous terms equal in the boundary-
layer equation.

\paragraph{Solution sketch.} In the boundary-layer momentum equation
the streamwise convective term scales as $u\,\partial u/\partial x
\sim U_\infty^2/x$ and the viscous term as $\nu\,\partial^2 u/
\partial y^2 \sim \nu U_\infty/\delta^2$. The boundary layer is the
region where these two balance, so $U_\infty^2/x \sim \nu U_\infty/
\delta^2$, giving $\delta^2 \sim \nu x/U_\infty$ and
$\delta \sim \sqrt{\nu x/U_\infty} = x/\sqrt{\text{Re}_x}$. The
order-one constant ($5.0$ for the $99\%$ thickness) comes from the
full Blasius solution, but the scaling is fixed by this balance.
\end{exercise}

\begin{exercise}
Derive the Kármán momentum integral relation $d\theta/dx = C_f/2$
(for zero pressure gradient) by integrating the boundary-layer
momentum equation across the layer.

\paragraph{Solution sketch.} Integrate the steady boundary-layer
momentum equation $u\,u_x + v\,u_y = \nu\,u_{yy}$ across the layer
from $y=0$ to $y\to\infty$, using continuity $u_x + v_y = 0$ to
eliminate $v$ and integrating by parts. With zero pressure gradient
and the no-slip and freestream boundary conditions, the left side
reduces to $\tfrac{d}{dx}\int_0^\infty u(U_\infty - u)\,dy =
U_\infty^2\,d\theta/dx$, where $\theta$ is the momentum thickness;
the right side is the wall shear $\tau_w/\rho = \nu\,(u_y)_{y=0}$.
Dividing by $U_\infty^2$ and using $C_f = \tau_w/(\tfrac12\rho
U_\infty^2)$ gives $d\theta/dx = C_f/2$. (With a pressure gradient
extra terms in $\delta^*$ and $dU_\infty/dx$ appear; this is the
zero-gradient special case.)
\end{exercise}

\begin{exercise}
Show that the Reynolds number is the only dimensionless parameter
governing the steady incompressible Navier-Stokes equation in a
geometry of fixed shape, by non-dimensionalisation with a
characteristic velocity and length.

\paragraph{Solution sketch.} Scale $\vec u$ by $U$, lengths by $L$,
pressure by $\rho U^2$, time by $L/U$. Substituting into the steady
incompressible Navier-Stokes equation $\rho(\vec u\cdot\nabla)\vec u
= -\nabla p + \mu\nabla^2\vec u$ and dividing through by $\rho U^2/L$
gives $(\vec u^*\cdot\nabla^*)\vec u^* = -\nabla^* p^* + (1/\text{Re})
\nabla^{*2}\vec u^*$ with $\text{Re} = \rho U L/\mu$. The only
parameter surviving the scaling is $\text{Re}$ (gravity, if retained,
would add $1/\text{Fr}^2$). The boundary conditions on a fixed-shape
geometry are themselves dimensionless once positions are scaled by
$L$, so two flows with the same shape and the same $\text{Re}$ have
identical non-dimensional solutions, which is the similarity
principle.
\end{exercise}

\begin{exercise}
Derive Stokes's drag law $F_D = 6\pi\mu a U$ from the creeping-
flow form of the Navier-Stokes equations around a sphere. Sketch
the derivation; the full algebra is in any standard reference.

\paragraph{Solution sketch.} At $\text{Re}\ll1$ drop the inertial
term, leaving the Stokes equations $\nabla p = \mu\nabla^2\vec u$,
$\nabla\cdot\vec u = 0$. Taking the divergence shows the pressure is
harmonic, $\nabla^2 p = 0$. With the boundary conditions $\vec u =
\vec 0$ on the sphere and $\vec u\to\vec U$ at infinity, the solution
is a Stokeslet plus a potential dipole; the resulting surface stress
integrates to a force along $\vec U$. Two-thirds of the force is
viscous skin friction and one-third is pressure drag, summing to
$F_D = 6\pi\mu a U$. The drag coefficient is then $C_D = F_D/(\tfrac12
\rho U^2\pi a^2) = 24/\text{Re}$ with $\text{Re} = \rho U(2a)/\mu$.
The full angular dependence of $\vec u$ and $p$ is in
\cite{paper:ch02t3b-stokes-1851}.
\end{exercise}

\subsection*{Estimation}\pathtag{core}

\begin{exercise}
Estimate the Reynolds number of human blood flow in the aorta.
State assumptions about blood density, viscosity, aortic diameter,
and mean flow velocity.

\paragraph{Reference estimate.} Take blood $\rho \approx 1060\,\si{
\kilogram\per\meter\cubed}$, $\mu \approx 3.5\times10^{-3}\,\si{
\pascal\second}$ (about $3.5\times$ water), aortic diameter $D
\approx 25\,\si{\milli\meter}$, mean velocity $\bar u\approx0.2\,\si{
\meter\per\second}$ (averaged over the cardiac cycle; peak systolic
is several times higher). Then $\text{Re}_D = \rho\bar u D/\mu =
1060\times0.2\times0.025/(3.5\times10^{-3}) \approx 1.5\times10^{3}$.
The mean-flow Reynolds number is just below the pipe-transition value,
so aortic flow is nominally laminar on average; peak-systolic
Reynolds numbers reach several thousand, producing transient,
disturbed flow near the valve, which is one reason aortic flow is
not cleanly laminar throughout the cycle.
\end{exercise}

\begin{exercise}
Estimate the boundary-layer thickness on a sailing yacht hull at
its trailing edge while sailing at $5\,\si{\meter\per\second}$.
State assumptions about hull length and water properties.

\paragraph{Reference estimate.} Take a waterline length $L\approx
10\,\si{\meter}$ and water $\nu\approx1.0\times10^{-6}\,\si{\meter}^2/
\si{\second}$. $\text{Re}_L = U L/\nu = 5\times10/(1.0\times10^{-6}) =
5\times10^{7}$, far above the transition value, so the layer is
turbulent over nearly the whole hull. Turbulent thickness at the
stern: $\delta = 0.37\,L\,\text{Re}_L^{-1/5} = 0.37\times10\times
(5\times10^7)^{-0.2} \approx 3.7\times0.0182 \approx 6.7\times10^{-2}
\,\si{\meter}$, about $7\,\si{\centi\meter}$. A boundary layer of
several centimetres at the stern is consistent with the observation
that the rearmost hull region sits in a thick, slow-moving sheath of
water.
\end{exercise}

\begin{exercise}
Estimate the drag force on a road cyclist at $10\,\si{\meter\per
\second}$ in still air. State assumptions about the cyclist's
frontal area, drag coefficient, and air density. Compare to a
typical cyclist's sustainable power output.

\paragraph{Reference estimate.} Take frontal area $A\approx0.4\,\si{
\meter}^2$, $C_D\approx0.9$ (upright rider plus bike), $\rho\approx
1.2\,\si{\kilogram\per\meter\cubed}$. Drag $F_D = \tfrac12\rho U^2 A
C_D = 0.5\times1.2\times10^2\times0.4\times0.9 \approx 21.6\,\si{
\newton}$. Drag power $P = F_D U \approx 216\,\si{\watt}$. A trained
amateur sustains roughly $200$ to $300\,\si{\watt}$, so $10\,\si{
\meter\per\second}$ ($36\,\text{km/h}$) is near the limit of what an
upright rider holds against air drag alone, which is why aerodynamic
position and clothing matter so much in time trials.
\end{exercise}

\begin{exercise}
Estimate the terminal velocity of a raindrop of $2\,\si{\milli
\meter}$ diameter falling in still air. State whether the relevant
drag regime is Stokes or beyond, and choose the appropriate
correlation.

\paragraph{Reference estimate.} A Stokes estimate, $U = (\rho_w-
\rho_a)g d^2/(18\mu) = 1000\times9.81\times(2\times10^{-3})^2/(18
\times1.8\times10^{-5}) \approx 121\,\si{\meter\per\second}$, gives a
Reynolds number $\text{Re} = \rho_a U d/\mu = 1.2\times121\times
0.002/(1.8\times10^{-5}) \approx 1.6\times10^{4}$, far outside Stokes
($\text{Re}<1$). The drag is in the constant-$C_D$ bluff-body range,
so balance weight against $F_D = \tfrac12\rho_a U^2 (\tfrac\pi4 d^2)
C_D$ with $C_D\approx0.5$: $U = \sqrt{\tfrac{4\rho_w g d}{3\rho_a
C_D}} = \sqrt{4\times1000\times9.81\times0.002/(3\times1.2\times0.5)}
\approx 6.6\,\si{\meter\per\second}$. (Real $2\,\si{\milli\meter}$
drops deform and fall near $6.5\,\si{\meter\per\second}$, so the
bluff-body estimate is the right one; Stokes overshoots by twenty
times.)
\end{exercise}

\begin{exercise}
Estimate the friction-related power dissipation in a $50\,\si{
\meter}$ section of household water-supply pipe of $20\,\si{\milli
\meter}$ diameter carrying $0.2\,\text{L/s}$. State the relevant
Reynolds number and friction factor.

\paragraph{Reference estimate.} $Q = 0.2\times10^{-3}\,\text{m}^3/
\text{s}$, $D = 0.02\,\si{\meter}$, area $= \pi D^2/4 = 3.14\times
10^{-4}\,\si{\meter}^2$, so $\bar u = Q/A \approx 0.64\,\si{\meter
\per\second}$. Water $\rho=998$, $\mu=10^{-3}$: $\text{Re}_D =
998\times0.64\times0.02/10^{-3} \approx 1.27\times10^{4}$, turbulent.
Smooth-pipe Haaland gives $f\approx0.029$. Pressure drop over
$L=50\,\si{\meter}$: $\Delta p = f\tfrac{L}{D}\tfrac{\rho\bar u^2}{2}
= 0.029\times2500\times998\times0.41/2 \approx 1.5\times10^{4}\,\si{
\pascal}$. Dissipated power $P = \Delta p\,Q = 1.5\times10^4\times
2\times10^{-4} \approx 3\,\si{\watt}$. A few watts lost to pipe
friction in a long household run is small but not negligible over a
building's lifetime of pumping.
\end{exercise}

\subsection*{Simulation}\pathtag{mastery}

\begin{exercise}
Implement a finite-difference solver for the boundary-layer
equations on a flat plate with zero pressure gradient. Verify the
Blasius profile by comparing the computed wall shear and
displacement thickness to the analytical values.

\paragraph{Solution sketch.} The cleanest route reuses the similarity
reduction: solve the Blasius ordinary differential equation $2f'''
+ ff'' = 0$ by shooting, as in \texttt{code/blasius.py}, and recover
$f''(0) = 0.33206$ (wall shear) and $\delta^*/\delta\approx0.345$
(displacement thickness). A direct marching finite-difference solver
of the boundary-layer partial differential equations discretises in
$y$, marches in $x$ with the freestream as the initial profile, and
should reproduce $C_f(x) = 0.664/\sqrt{\text{Re}_x}$ to within the
grid-refinement error; a wall-normal spacing fine near $y=0$ is
needed to resolve the steep wall gradient. Verification is a
convergence study: halve $\Delta y$ and confirm $f''(0)$ approaches
$0.332$ at the expected order.
\end{exercise}

\begin{exercise}
Implement a panel method for two-dimensional incompressible viscous
flow past a circular cylinder, including a thin-boundary-layer
correction (Thwaites' method for laminar, Head's method for
turbulent). Verify against published $C_D$ vs.\ $\text{Re}$ data.

\paragraph{Solution sketch.} Build the inviscid potential flow from a
distribution of source or vortex panels on the cylinder, giving the
pressure distribution $C_p(\theta)$. Feed the resulting edge velocity
into a boundary-layer integral method (Thwaites for the laminar
portion, Head for the turbulent portion after a transition criterion)
to locate the separation point where the wall shear vanishes.
Pressure drag follows from integrating $C_p$ over the attached
surface up to separation; skin-friction drag follows from the
boundary-layer shear. Validate against
\texttt{data/sphere\_drag\_curve.csv} (or a cylinder equivalent): the
method should reproduce the subcritical plateau and, with a tripped
transition, the drag-crisis drop. A panel method without the
boundary-layer coupling returns the d'Alembert zero drag, which is
the negative control that shows why the viscous correction is the
whole point.
\end{exercise}

\begin{exercise}
Run a simple LES of channel flow at $\text{Re}_\tau \approx 200$
using an existing open-source CFD code. Compare the mean velocity
profile in the log layer to the law-of-the-wall prediction
$u^+ = (1/\kappa)\ln y^+ + B$.

\paragraph{Solution sketch.} Set up a periodic channel at friction
Reynolds number $\text{Re}_\tau = u_\tau h/\nu\approx200$ with a
sub-grid-scale model (Smagorinsky or dynamic), driven by a constant
mean pressure gradient that fixes $u_\tau$. Run until the statistics
are stationary, then time-average the streamwise velocity, scale by
$u_\tau$ and $\nu$ to wall units, and overplot against
figure~\ref{fig:vol03ch12:law-of-wall}. The mean profile should show
$u^+ = y^+$ in the viscous sublayer and $u^+\approx(1/0.41)\ln y^+ +
5.0$ in the log region; the verification check is that the constants
$\kappa$ and $B$ come out near the standard values and that the
wall-normal resolution places several points inside $y^+<5$. A
coarse grid that misses the sublayer over-predicts the wall velocity
and is the diagnostic for insufficient near-wall resolution.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a piping system for a building's hot-water supply,
delivering $0.5\,\text{L/s}$ to each of $20$ fixtures across a
$30\,\si{\meter}$ run. Specify the pipe diameter at the trunk and
at the branches, and check the pressure drop against a maximum
allowable of $50\,\text{kPa}$.

\paragraph{Model-answer sketch.} Trunk flow if all fixtures draw at
once is $20\times0.5 = 10\,\text{L/s}$, but a diversity factor of
$0.3$ to $0.5$ is standard, so size the trunk for roughly $4\,\text{
L/s}$. Pick a target velocity near $1.5\,\si{\meter\per\second}$ to
limit noise and erosion: $A = Q/\bar u = 4\times10^{-3}/1.5 =
2.7\times10^{-3}\,\si{\meter}^2$, so $D\approx0.058\,\si{\meter}$;
choose a $54\,\si{\milli\meter}$ or $63\,\si{\milli\meter}$ standard
size. Branches carry $0.5\,\text{L/s}$ at the same velocity:
$D\approx0.02\,\si{\meter}$ ($20\,\si{\milli\meter}$). Check the
trunk drop with \texttt{code/pipe\_pressure\_drop.py}:
$\text{Re}_D = \rho\bar u D/\mu\approx
998\times1.5\times0.06/10^{-3}\approx9\times10^{4}$, turbulent,
$f\approx0.02$, $\Delta p = f(L/D)\tfrac{\rho\bar u^2}{2}\approx
0.02\times(30/0.06)\times998\times1.5^2/2\approx1.1\times10^{4}\,\si{
\pascal}$, well under $50\,\text{kPa}$. Add fitting and elevation
losses; if the margin is comfortable, a smaller trunk may be
acceptable. The deliverable states diameters, velocities, the
computed drop, and the margin to the limit.
\end{exercise}

\begin{exercise}
Design a settling tank for separating sand ($\rho_s = 2650\,\si{
\kilogram\per\meter\cubed}$) from a water flow at $0.01\,\text{m}
^3\text{/s}$. Specify the tank's depth, length, and width, given
that all particles larger than $0.5\,\si{\milli\meter}$ must
settle out before exiting.

\paragraph{Model-answer sketch.} The settling velocity of a
$0.5\,\si{\milli\meter}$ sand grain sets the design. A Stokes
estimate gives $U_s = (\rho_s-\rho_f)g d^2/(18\mu) = 1650\times9.81
\times(5\times10^{-4})^2/(18\times10^{-3})\approx0.22\,\si{\meter\per
\second}$, but $\text{Re}=\rho_f U_s d/\mu\approx110$ is well past
Stokes, so use the intermediate-range drag and the iterative solver,
which lowers $U_s$ to roughly $0.07\,\si{\meter\per\second}$. The
ideal-settling criterion is that the overflow rate (flow divided by
plan area) not exceed $U_s$: $Q/(LW)\le U_s$, so $LW\ge0.01/0.07
\approx0.14\,\si{\meter}^2$. Pick depth so the horizontal residence
time exceeds the settling time from the surface: with depth $H$,
$t_{\text{settle}}=H/U_s$ and $t_{\text{residence}}=L W H/Q$, which
reduces to the same plan-area criterion. Choose, for example,
$L=2\,\si{\meter}$, $W=0.5\,\si{\meter}$ (plan area $1\,\si{\meter}
^2$, comfortably above the minimum), depth $H\approx0.5\,\si{\meter}$,
and keep the inlet velocity low to avoid re-suspension. The
deliverable states the settling velocity with its regime check, the
plan-area criterion, and the chosen dimensions with margin.
\end{exercise}

\begin{exercise}
Design an exhaust stack of $50\,\si{\meter}$ height and $3\,\si{
\meter}$ diameter to survive a $25\,\si{\meter\per\second}$ wind
without vortex-induced resonance. Estimate the Strouhal frequency
and the stack's fundamental natural frequency; specify any
aerodynamic spoilers or structural changes needed.

\paragraph{Model-answer sketch.} Shedding frequency at the design
wind: $\text{Re} = UD/\nu = 25\times3/(1.5\times10^{-5}) =
5\times10^{6}$, so $\text{St}\approx0.2$ and $f_s = \text{St}\,U/D =
0.2\times25/3\approx1.7\,\si{\hertz}$. A $50\,\si{\meter}$ cantilever
stack has a low fundamental, on the order of $0.3$ to $1\,\si{\hertz}$
depending on wall thickness and base fixity; estimate it from
$f_n\approx(1.875)^2/(2\pi)\sqrt{EI/(\rho_L L^4)}$ for a uniform
cantilever. Lock-in occurs at $U_{\text{lock}} = f_n D/\text{St}$; if
the fundamental is near $0.5\,\si{\hertz}$, lock-in falls at $U
\approx7.5\,\si{\meter\per\second}$, a frequent wind, so vortex-
induced vibration is a real risk across the operating range. Remedies:
helical strakes over the top third of the stack (the standard fix,
disrupting coherent shedding), a tuned mass damper, or stiffening to
move $f_n$ out of the common-wind lock-in band. The deliverable
states $f_s$, the estimated $f_n$, the lock-in speed, and the chosen
countermeasure with its rationale.
\end{exercise}

\begin{exercise}
Design the underbody and rear-end shape of a road vehicle to
reduce its drag coefficient from a baseline $C_D = 0.35$ to a
target $0.28$. State the principal mechanisms (rear-spoiler shape,
underbody flatness, wheel housing, trailing-edge taper) and
estimate the contribution of each.

\paragraph{Model-answer sketch.} The target is a $20\%$ reduction in
$C_D$, of which the largest single share usually comes from managing
rear separation: a tapered rear (boat-tail) or a properly sized
spoiler delays or organises separation and narrows the wake, worth
perhaps $0.03$ to $0.04$ in $C_D$. A flat, sealed underbody with a
rear diffuser reduces underbody turbulence and pressure drag, worth
another $0.01$ to $0.02$. Wheel-housing treatment and partial wheel
shrouding cut the disturbed flow around the wheels, worth around
$0.01$. Smoothing the front (rounded leading edges, managed cooling-
air path) trims a further $0.005$ to $0.01$. Summing the central
estimates gives roughly $0.06$ to $0.08$, enough to move $0.35$ to
$0.28$ if the measures combine without too much interaction. The
deliverable lists each mechanism, its estimated contribution, and the
caveat that the contributions are not strictly additive and need
wind-tunnel or CFD confirmation.
\end{exercise}

\subsection*{Diagnosis and reverse engineering}\pathtag{core}

\begin{exercise}
A pump is delivering only $60\,\%$ of its rated flow at the rated
discharge pressure. Identify three viscous-flow mechanisms (pipe
fouling, partial valve closure, cavitation in the pump) that could
produce this and the diagnostic that would distinguish among them.

\paragraph{Solution sketch.} Fouling raises the effective roughness
and lowers the diameter, shifting the system curve up; the diagnostic
is to inspect or borescope the line and to compare the measured
pressure drop at known flow with the new-pipe Moody prediction (a
drop larger than predicted points to fouling). A partially closed
valve adds a localised loss; the diagnostic is to check valve
position and to look for a large, concentrated pressure drop across
the valve rather than distributed along the run. Cavitation in the
pump shows as a flow-and-head deficit accompanied by noise and
vibration, worsening as suction pressure falls; the diagnostic is to
raise the net positive suction head (lower the pump, raise the supply
level, or throttle the discharge rather than the suction) and see
whether the deficit clears. The three are distinguished by where the
loss is concentrated (distributed, at the valve, or at the pump) and
by the acoustic and suction-pressure signatures.
\end{exercise}

\begin{exercise}
An exhaust stack on a power station has begun to oscillate
visibly under moderate winds. Identify the most probable
mechanism, the test that would confirm it, and one immediate and
one permanent remedial measure.

\paragraph{Solution sketch.} The most probable mechanism is vortex-
induced vibration: at the prevailing wind speed the Strouhal shedding
frequency has come into lock-in with one of the stack's natural
frequencies. The confirming test is to measure the oscillation
frequency and the wind speed, compute $f_s = \text{St}\,U/D$, and
check that the observed motion frequency matches both $f_s$ and a
modal frequency of the stack (an instrumented modal survey settles
it). Immediate measure: restrict operation or temporarily attach
guy cables or a perimeter rope-and-chain spoiler to break the
shedding coherence. Permanent measure: fit helical strakes over the
upper third of the stack, or install a tuned mass damper, or stiffen
the structure to move $f_n$ out of the common-wind lock-in band.
\end{exercise}

\begin{exercise}
A wind-tunnel test of a passenger-car model at $1/5$ scale gives
$C_D = 0.42$, but full-scale road testing gives $C_D = 0.32$.
Identify the most probable cause of the discrepancy, given that
the wind tunnel was operated at a freestream velocity matching
the full-scale highway speed.

\paragraph{Solution sketch.} The probable cause is a Reynolds-number
mismatch. At $1/5$ scale with the same speed, the model Reynolds
number is one-fifth of the full-scale value, $\text{Re}_m =
\text{Re}_p/5$. If that lower Reynolds number leaves the model in a
subcritical regime (laminar separation, wider wake) while the full-
scale vehicle runs supercritical (turbulent boundary layer, delayed
separation, narrower wake), the model reads a higher $C_D$. The fix
for similarity is to match Reynolds number, which would require a
tunnel speed of five times the road speed (often impractical) or a
pressurised or cryogenic tunnel; the practical fix is to trip the
model boundary layer to turbulence with roughness strips so it
separates like the full-scale flow. The discrepancy is the everyday
reason model drag must be Reynolds-corrected before it is trusted.
\end{exercise}

\begin{exercise}
A water-supply pipe section has measured pressure drop $20\,\%$
larger than the Moody-chart prediction for new pipe of its
material. Identify what physical change in the pipe could account
for this and the inspection that would confirm or refute it.

\paragraph{Solution sketch.} A $20\%$ excess pressure drop at fixed
flow points to increased roughness or reduced bore: scale or
corrosion deposits raise $\epsilon/D$ and shrink $D$, both of which
raise the friction factor (and the $D^{-5}$ sensitivity of
Darcy-Weisbach makes even a small bore reduction strong). Back out the
implied roughness: solve Colebrook for the $\epsilon/D$ that
reproduces the measured $f$ at the operating Reynolds number, and
compare to the new-pipe handbook value. The confirming inspection is
a borescope or an ultrasonic wall-thickness and deposit survey of the
inner wall; a cut sample settles it. If the bore and wall are clean,
revisit the flow-rate measurement and the temperature (hence
viscosity) before blaming the pipe.
\end{exercise}

\subsection*{Failure analysis}\pathtag{core}

\begin{exercise}
The BEA report on Air France Flight 447 \cite{acc:bea-af447}
identifies a sequence beginning with pitot-tube icing and ending
with deep stall and impact. Argue, in $200$ to $300$ words, how
the boundary-layer physics of high-altitude stall constrains the
recovery procedure, and what design discipline (cockpit
indication, autoflight handoff, crew training) the case
illuminates.

\paragraph{Solution sketch.} A strong answer ties the recovery
constraint to the separated-flow physics: in a deep stall the wing
boundary layer is massively separated, lift has collapsed and drag is
several times its attached value, so the only way to reattach the
flow is to lower the angle of attack, which at altitude demands a
nose-down attitude and a large loss of altitude to rebuild airspeed.
The thin air at $35{,}000\,\text{ft}$ leaves little margin between
stall and overspeed, and the separated wake degrades control
authority. The case illuminates three disciplines: unambiguous
cockpit indication of the stalled state (angle-of-attack display,
stall warning that stays valid under unreliable airspeed), clean
autoflight handoff that does not drop the crew into manual control at
a high workload moment, and crew training in high-altitude upset and
unreliable-airspeed recovery \cite{acc:bea-af447}. The
boundary-layer point is that stall is a flight-envelope boundary set
by separation, not a recoverable nuisance.
\end{exercise}

\begin{exercise}
A diffuser at the inlet of a centrifugal compressor stage stalls
intermittently at off-design flow rates, producing surge cycles.
Identify the boundary-layer mechanism, the parameter (typically
diffuser flow angle or diffuser area ratio) that controls it, and
the design countermeasure.

\paragraph{Solution sketch.} The mechanism is boundary-layer
separation in the diffuser (vaned or vaneless) under the adverse
pressure gradient that the diffuser imposes to recover dynamic
pressure. At off-design flow the incidence onto the diffuser changes,
the local pressure gradient steepens, the boundary layer separates,
and the stage cannot sustain the pressure rise, which couples into the
cyclic charge-and-discharge of surge. The controlling parameter is the
diffuser inlet flow angle (equivalently the area ratio and divergence),
which sets whether the gradient stays within the attached limit.
Countermeasures: variable-geometry diffuser vanes that re-aim with
flow, bleed or recirculation to widen the stable range, and a surge
or stall margin built into the operating line so the machine does not
run near the separation boundary.
\end{exercise}

\begin{exercise}
A modern long-span suspension bridge has been instrumented with
accelerometers and has shown low-amplitude oscillation at
specific moderate wind speeds. Argue, in $200$ to $300$ words,
the boundary-layer / vortex-shedding mechanism most likely
responsible, distinguishing it from the larger-amplitude
aeroelastic flutter that destroyed the original Tacoma Narrows
bridge \cite{acc:tacoma-narrows-1941}.

\paragraph{Solution sketch.} The low-amplitude, narrow-wind-speed-band
oscillation is the signature of vortex-induced vibration: the Kármán
shedding frequency $f_s = \text{St}\,U/D$ has come into lock-in with a
deck or cable mode, and the response self-limits because the shedding
forcing is bounded and the lock-in range is narrow. The mechanism is
forced resonance by a bounded aerodynamic load. Torsional flutter, the
larger-amplitude phenomenon that destroyed the original Tacoma Narrows
deck, is different in kind: it is a self-excited aeroelastic
instability in which the deck motion feeds energy into itself, so the
amplitude grows without bound above a critical wind speed rather than
locking to a shedding frequency. The distinguishing observations are
amplitude (bounded versus divergent), the wind-speed dependence (a
narrow lock-in band versus a critical-speed threshold above which it
grows), and whether the motion frequency tracks $\text{St}\,U/D$
(vortex shedding) or sits at a structural torsional mode independent
of $U$ (flutter).
\end{exercise}

\subsection*{Judgment}\pathtag{standard}

\begin{exercise}
A senior engineer argues that pipe-flow design can be done with
the Hagen-Poiseuille relation across the full operating range,
because the typical design Reynolds number is well below $2300$.
Write a one-paragraph reply that takes the argument seriously,
identifies the operating-range conditions under which it is true,
and identifies the conditions under which it fails.

\paragraph{Solution sketch.} A fair reply grants the case where it
holds: for genuinely viscous-dominated flows (very low velocity, small
bore, or high-viscosity fluid) the Reynolds number stays below $2300$
across the operating range, and Hagen-Poiseuille is then exact, not
approximate. The argument fails wherever the operating range reaches
the transition: most water and air distribution systems run at
$\text{Re}_D$ of $10^4$ to $10^6$, where the flow is turbulent and the
friction factor follows the Moody chart, not $64/\text{Re}_D$. The
test is to compute $\text{Re}_D$ at the highest design flow, not the
nominal one; a system that is laminar at average flow can be turbulent
at peak draw. The honest position is that Hagen-Poiseuille is the right
tool only after the regime check confirms laminar flow at the worst-
case operating point.
\end{exercise}

\begin{exercise}
A junior engineer asks why CFD results are reported with three
significant figures when the underlying turbulence closure
introduces uncertainties of $10$ to $30\,\%$. Write a one-paragraph
reply that distinguishes verification from validation and the
status of CFD predictions absent experimental anchoring.

\paragraph{Solution sketch.} The reply separates two things. The three
significant figures report the precision with which the code solved its
discretised equations, which is a verification question answered by
mesh refinement and order-of-accuracy checks; a converged solver can
report many digits. Those digits say nothing about whether the
equations and the turbulence closure model the real flow, which is the
validation question, answered only by comparison with experiment. A
closure carrying $10$ to $30\%$ model uncertainty means the prediction
is accurate to that band regardless of how many digits the solver
prints. The honest report states the converged value with its
discretisation error and, separately, the model-form uncertainty from
the closure; an unvalidated CFD result is a precise solution of an
approximate model, not a prediction of the flow.
\end{exercise}

\begin{exercise}
A designer proposes to omit boundary-layer-transition analysis on
a wind-turbine blade because the operating Reynolds number is in
the fully turbulent range. Write a one-paragraph response
addressing whether transition still matters and, if so, on what
parts of the blade.

\paragraph{Solution sketch.} Transition still matters. The chord-based
Reynolds number on a wind-turbine blade falls sharply toward the root,
where the chord and local speed are small, so the inboard sections can
sit in a transitional or even laminar regime even when the tip is
fully turbulent. The blade also carries a laminar run near the leading
edge before transition, and the transition location moves with angle
of attack and with surface contamination (insect debris, erosion,
rain), which shifts the drag and the separation behaviour. The
practical consequences are a measurable drag and annual-energy penalty
from leading-edge roughness and a sensitivity of stall behaviour to
transition. The response is that transition analysis is warranted on
the inboard sections and on the leading-edge region, and that surface
cleanliness and erosion protection are part of keeping the assumed
transition behaviour valid.
\end{exercise}

\begin{exercise}
Identify a viscous-flow situation in your immediate environment
where the boundary-layer regime controls observable behaviour.
State, in $200$ to $300$ words, the Reynolds number you estimate,
the regime it corresponds to, the dominant viscous mechanism, and
the consequence you observe.

\paragraph{Solution sketch.} Open-ended; a complete answer names a
concrete situation, estimates its Reynolds number with stated length
and velocity scales, places it in a regime, names the mechanism, and
ties it to an observation. One worked example: honey dripping from a
spoon, length scale $\sim10\,\si{\milli\meter}$, velocity $\sim
0.1\,\si{\meter\per\second}$, $\nu\sim10^{-2}\,\si{\meter}^2/\si{
\second}$, so $\text{Re}\sim0.1\ll1$, the creeping-flow regime
dominated by viscosity, which is why the thread is smooth and slow and
does not break into droplets the way water does. Another: cigarette
smoke rising in still air, $\text{Re}$ of a few hundred near the
source, where the laminar plume is visibly smooth before it transitions
to a turbulent, chaotically mixing column a short distance up, a
direct sighting of the laminar-to-turbulent transition. The grading
point is that the estimate, the regime, the mechanism, and the
observation hang together.
\end{exercise}
