\chapter{The discipline of estimation}
\label{vol01:ch09}

\epigraph{Almost any physical quantity worth knowing can be estimated to
within a factor of three before any calculation is done; once the
calculation is done, the estimate is what tells you whether the
calculation is sane.}{The Fermi habit, as recorded in the working
practice of many engineers and stated formally in Mahajan's \emph{The Art
of Insight} \cite{text:mahajan2014}.}

\chapmeta{Half-life: foundational. Archetypes: scaling (canonical home of
order-of-magnitude reasoning), uncertainty. Exercise target: 50. Project
track: analysis only. Hazard class: none. Reader path default: standard,
with sections explicitly tagged. Named cases: Space Shuttle Challenger
(1986).}

\noindent
We close Volume I by codifying a habit the reader has been practising for
eight chapters. Estimation is the engineer's first tool, used before the
calculator and the spreadsheet, and the last check before a number leaves
the desk. We put the moves in one place, name the \verb|estimation|
environment as their formal home, and then make the reader use them on
fifty problems across physical, urban, biological, technological, and
environmental scales.

The fifty Fermi problems in section~9.9 are also the volume's closing
project. Per the editorial settlement on this chapter's project track,
the project and the exercise set are the same artifact: solutions to all
fifty problems with the reader's working written out, comparison to
ground truth where it can be looked up, and a 1500-word reflection on
which problem types the reader's intuition was best at and worst at.
The deliverable is the reader's evidence that the Volume I habit has
become usable.

\section{Order-of-magnitude reasoning}\pathtag{core}

An order of magnitude is a factor of ten. To reason at order-of-magnitude
precision is to ask, of every quantity, only which power of ten it
sits closest to. The mass of an apple is $10^{-1} \,\si{\kilogram}$. The
mass of a person is $10^{2} \,\si{\kilogram}$. The mass of a car is
$10^{3} \,\si{\kilogram}$. The mass of a long-haul aircraft at maximum
take-off weight is $10^{5} \,\si{\kilogram}$. The mass of a fully laden
container ship is $10^{8} \,\si{\kilogram}$. Six powers of ten separate
the apple from the ship, and an engineer who carries those six numbers in
working memory rarely confuses the kinds of system the numbers describe.

\Cref{fig:vol01:ch09:scale-ladder} extends the table across forty
decades of length and over fifty decades of mass. The figure is one
page because the order-of-magnitude ladder is the chapter's central
mnemonic; a reader returning to the ladder during later volumes
needs every benchmark in one view.

\input{volumes/01-quantity/ch09-discipline-of-estimation/figures/fig-scale-ladder.tex}

The case for working at this resolution is practical. A spreadsheet error
of a factor of two is invisible to the eye; an error of a factor of ten
is sometimes invisible too, when the answer falls in a familiar range. An
error of a factor of $10^{4}$ stands up off the page once the engineer
has the order-of-magnitude scale of the answer in mind before reading
it. Estimation supplies that scale. The reader who has estimated the
mass of the aircraft as $10^{5} \,\si{\kilogram}$ before opening the
manufacturer's reference card cannot misread $396{,}890 \,\si{\kilogram}$
as $39.6 \,\si{\kilogram}$ or $3.96 \times 10^{8} \,\si{\kilogram}$. The
estimate has narrowed the space of plausible readings before the reading
is taken.

\subsection{The internal ruler}

Volume I Chapter 1 introduced the engineer's \engterm{internal ruler}: a
mental table that maps quantities in a domain to their order of
magnitude. We can now describe how the table is built. The reader
acquires it by repeated practice on quantities of varying scale. Each
estimate, even a wrong one, leaves a benchmark in memory. After several
hundred estimates the reader can place a new quantity within an order of
magnitude without any conscious arithmetic, by interpolation between the
benchmarks already laid down.

The ruler is domain-specific. A practising mechanical engineer carries
masses, lengths, forces, pressures, and torques to within an order of
magnitude across the working range of their domain. A practising
electrical engineer carries voltages, currents, capacitances,
resistances, and frequencies. A practising civil engineer carries
volumes of concrete, costs per cubic metre, allowable bearing pressures,
and span-to-depth ratios. Crossing into a new domain is the engineer's
recurring challenge, because the existing ruler does not transfer; it has
to be rebuilt from a small number of anchor points and from repeated
estimation against ground truth.

The book's twelve volumes hand the reader anchor points for each domain.
Volume I gives metrology and statistics; Volume II gives mathematical
operations and computational complexity; Volume III gives forces,
energies, and powers; Volume IV gives flows, transport, and waves; and so
on. Each volume's estimation exercises serve to embed the anchor points
through use rather than through memorisation.

\subsection{The factor-of-three convention}

Engineering estimation by convention aims for a result within a factor
of three of the truth, in either direction. The convention is loose
enough to be reachable by mental arithmetic and tight enough to be
useful. A factor-of-three estimate distinguishes a process that is
feasible from one that is impossible, and a budget that is plausible
from one that is fictitious. It does not distinguish a process that is
elegant from one that is wasteful; that is what calculation is for.

The convention has structural roots, but the structure is logarithmic.
If $X = \prod_{i} f_{i}$, then multiplicative errors add in $\ln X$.
Four independent factor errors with log-width $\ln(\sqrt{3})$ combine in
quadrature to a log-width of $2\ln(\sqrt{3}) = \ln 3$, which is a factor
of $3$. Worst-case multiplication is looser: four factors each high by
$\sqrt{3}$ would put the product high by $(\sqrt{3})^{4} = 9$. The
factor-of-three target is therefore a working statistical target for
independent factor errors, not a worst-case guarantee.

Where the underlying physics demands tighter control, estimation moves
to a factor-of-two or factor-of-1.5 target. Where the underlying physics
permits looser control, the engineer reaches for an order-of-magnitude
estimate, accepting a factor of ten. The match between the convention
and the application is itself part of the discipline.

\subsection{The working anchor table}

The factor-of-three convention is reachable only if the engineer
already carries a small set of memorised order-of-magnitude
anchors. The table below is the working aide-memoire from which
most of the chapter's decompositions start. The reader who has
not yet memorised the table copies it onto a single index card
and carries the card until the values become automatic; the
working form is then no longer a card but a recall faster than
the question.

The atomic and physical-constant anchors fix the scale of any
microscopic estimate: Avogadro's number is $N_{A} = 6.02 \times
10^{23} \,\text{mol}^{-1}$, the atomic mass unit is $1 \,\text{u}
= 1.66 \times 10^{-27} \,\si{\kilogram}$, the Planck constant is
$h = 6.63 \times 10^{-34} \,\si{\joule\second}$, the Boltzmann
constant is $k_{B} = 1.38 \times 10^{-23} \,\si{\joule\per\kelvin}$,
the gas constant is $R = 8.314 \,\si{\joule\per\mole\per\kelvin}$,
the elementary charge is $e = 1.602 \times 10^{-19} \,
\si{\coulomb}$, and the speed of light in vacuum is $c = 2.998
\times 10^{8} \,\si{\meter\per\second}$. The $2019$ SI redefinition
fixed each of these as an exact-by-definition constant
\cite{web:ch09t3b-codata-2018}. The working engineer carries each
to two or three significant figures and looks up the rest only
when the calculation's stated precision requires them.

The geophysical anchors fix the scale of any planetary or
infrastructural estimate. The Earth's mean radius is
$R_{\oplus} = 6{,}371 \,\si{\kilo\meter}$; the Earth's mass is
$M_{\oplus} = 5.97 \times 10^{24} \,\si{\kilogram}$; the
Earth-Sun mean distance (one astronomical unit) is
$1.496 \times 10^{11} \,\si{\meter}$; the solar irradiance at the
top of the atmosphere is $S_{0} = 1{,}361 \,\si{\watt\per\meter\squared}$
\cite{data:ch09t3b-iso-21348-2007}; the atmospheric pressure at sea
level is $p_{0} = 101.325 \,\si{\kilo\pascal} = 1 \,\text{atm}$;
the gravitational acceleration is $g = 9.81 \,
\si{\meter\per\second\squared}$; the speed of sound in dry air at
$20\,\degC$ is $343 \,\si{\meter\per\second}$ and in seawater at
$25\,\degC$ is $\approx 1{,}500 \,\si{\meter\per\second}$
\cite{web:ch09t3b-nist-rp-2024}; and the density of water at
$4\,\degC$ is $1{,}000 \,\si{\kilogram\per\meter\cubed}$.

The bioenergetic anchors fix the scale of any human-system
estimate. A resting adult dissipates $\approx 100 \,\si{\watt}$
(basal metabolic rate of about $1{,}800 \,\text{kcal/day}$,
$= 1{,}800 \times 4{,}184 / 86{,}400 \,\si{\joule\per\second}
\approx 87 \,\si{\watt}$, rounded to $100 \,\si{\watt}$ for
working use); an active adult during
sustained moderate exercise dissipates $\approx 300 \,\si{\watt}$;
an elite cyclist sustains $\approx 400 \,\si{\watt}$ for an hour
and $\approx 1{,}500 \,\si{\watt}$ in a sprint. A heart's
mechanical power output is $\approx 1.3 \,\si{\watt}$ at rest. The
human eye's photopic response peaks at $\lambda = 555 \,
\si{\nano\meter}$ and the audible range spans $20 \,\si{\hertz}$
to $20 \,\si{\kilo\hertz}$ in young listeners.

The energy-density anchors fix the scale of any thermal or
combustion estimate. Gasoline's specific energy is $47 \,
\si{\mega\joule\per\kilogram}$ (lower heating value $42$ to $44
\,\si{\mega\joule\per\kilogram}$, working figure $45 \,
\si{\mega\joule\per\kilogram}$); diesel is $43 \,
\si{\mega\joule\per\kilogram}$; bituminous coal is $24 \,
\si{\mega\joule\per\kilogram}$; natural gas is $50 \,
\si{\mega\joule\per\kilogram}$; dry hardwood is $18 \,
\si{\mega\joule\per\kilogram}$. Lithium-ion pack specific energy
is $\approx 250 \,\si{\watt\hour\per\kilogram} \approx 0.9 \,
\si{\mega\joule\per\kilogram}$ in $2024$ production cells, a
factor of $50$ below gasoline. Uranium-$235$ fission releases
$\approx 80 \,\si{\tera\joule\per\kilogram}$, six orders of
magnitude above any chemical fuel
\cite{data:ch09t3b-energy-densities-2023}.

The demographic and economic anchors fix the scale of any
aggregate human estimate. The world population is $8.1 \times
10^{9}$ as of mid-$2024$ \cite{data:ch09t3b-un-wpp-2024}; the
United States population is $3.34 \times 10^{8}$; the European
Union population is $4.5 \times 10^{8}$; the Chinese population
is $1.41 \times 10^{9}$; the Indian population is $1.43 \times
10^{9}$. The world's mean energy consumption per person is
$\approx 75 \,\si{\giga\joule}/\text{yr}$ ($\approx 2{,}400 \,
\si{\watt}$ continuous); the US figure is $\approx 280 \,
\si{\giga\joule}/\text{yr}$ ($\approx 9{,}000 \,\si{\watt}$)
\cite{data:ch09t3b-iea-wer-2024}. World gross domestic product
was $\approx 1.05 \times 10^{14} \,\text{USD}$ (current US dollars,
current as of $2023$) \cite{data:ch09t3b-wb-gdp-2024}; global
internet protocol traffic was $\approx 4.8 \,
\si{\zetta\byte}/\text{yr}$ (current as of $2023$)
\cite{data:ch09t3b-cisco-visual-2020}, or roughly $150 \,
\si{\tera\byte\per\second}$ averaged across the year.

The conversion ratios that the engineer carries in working memory
close the table. One horsepower is $745.7 \,\si{\watt}$; one
calorie is $4.184 \,\si{\joule}$; one British thermal unit is
$1{,}055 \,\si{\joule}$; one electron-volt is $1.602 \times
10^{-19} \,\si{\joule}$; one kilowatt-hour is $3.6 \,
\si{\mega\joule}$; one therm is $105.5 \,\si{\mega\joule}$. One
mile is $1.609 \,\si{\kilo\meter}$; one foot is $0.3048 \,
\si{\meter}$; one nautical mile is $1.852 \,\si{\kilo\meter}$. One
pound-force is $4.448 \,\si{\newton}$; one pound-mass is $0.4536
\,\si{\kilogram}$; one atmosphere is $101.325 \,
\si{\kilo\pascal} = 14.696 \,\text{psi}$; one bar is exactly
$10^{5} \,\si{\pascal}$; one millimetre of mercury is $133.3 \,
\si{\pascal}$.

The fifty-anchor memorised set is the working capital from which
any Fermi problem can be opened. A reader who has the table at
recall can place a new estimate within an order of magnitude in
seconds; a reader who must look up each anchor takes minutes, and
the discipline degrades into a research task. The chapter's fifty
problems at section~9.10 cover most of the table's domains; the
reader who completes the fifty has, as a byproduct, memorised the
anchors.

\subsection{The first canonical estimate}

The chapter's first formal estimation block trains the basic moves. The
question has an external comparison value, but the work is in the
decomposition, the recorded assumptions, and the postmortem.

\begin{estimation}
\textbf{How many piano tuners work in Chicago?}

\smallskip
\textit{Estimate, before reading on. The classic Fermi problem.}

\smallskip
The estimate proceeds by decomposition into factors that the reader can
each estimate within a factor of two or three. The product of the
factors is the estimate.

The Chicago-Naperville-Elgin metropolitan area had a population of
$9.41 \times 10^{6}$ on 1 July 2024, in the U.S. Census Bureau Vintage
2024 estimate \cite{data:census-cbsa-2024}. We round that to $9 \times
10^{6}$ for the estimate and take it as the denominator.

How many of those people own a piano in their household? The fraction is
small. Pianos are expensive, take floor space, and require a household
that values an acoustic instrument. Call it $1$ piano per $40$
households, with $2.5$ people per household; that is one piano per
$100$ people. The number of pianos in Chicago is then approximately
$9 \times 10^{4}$.

How often is a piano tuned? A serious player tunes once or twice a year;
an unused household piano is tuned once every several years; a heavily
played concert instrument is tuned weekly. Averaged across the
population of pianos in a city, call the rate one tuning per piano per
year. The annual demand for tunings is then about $9 \times 10^{4}$.

How many tunings does one tuner perform per year? A tuning takes about
two hours in the home plus travel; a working tuner can perform two or
three per day. Calling it $750$ tunings per year (about three per
working day, $250$ working days), one tuner supplies about $750$
tunings.

The number of tuners is then about $9 \times 10^{4} / 750 \approx 120$.
The estimate carries an uncertainty of perhaps a factor of three in
either direction, dominated by the fraction-of-households-with-pianos
factor and by the average-tunings-per-piano factor. The defensible
range is roughly $40$ to $400$.

For comparison, the U.S.~Bureau of Labor Statistics OEWS table for May
2023 reports $350$ musical instrument repairers and tuners in the
Chicago-Naperville-Elgin metropolitan area, occupation code 49-9063,
with a relative standard error of $24.6\,\%$
\cite{data:bls-oews-2023-chicago}. That is within our defensible range.

The postmortem matters more than the answer. Two of the four factors
were within a factor of two of the truth. The household-with-piano
fraction was the dominant uncertainty; a reader with direct knowledge
of musical-instrument ownership rates in Chicago could narrow the
estimate by a factor of two. The general lesson: each factor in a
Fermi decomposition carries its own uncertainty; the dominant one sets
the tightness of the result.
\end{estimation}

This estimate has done four things at once. It decomposed the question
into factors. It estimated each factor within a factor of two or three
using only working knowledge. It multiplied to a result. It compared
the result to a known reference and identified the factor that
dominated the uncertainty. Those four moves recur in every Fermi
estimate the rest of the chapter trains.

\Cref{fig:vol01:ch09:fermi-tree} draws the same decomposition as a
tree. The root carries the answer; the four boxes at depth one are
the four factors written out above; two of them, the pianos-per-person
and the tuner-throughput-per-year factors, expand at depth two into
sub-factors that the reader's working knowledge can each defend on
their own. A reference implementation of the same calculation, with
the low, nominal, and high inputs explicit, is in
\repopath{volumes/01-quantity/ch09-discipline-of-estimation/code/piano_tuners.py}.

\input{volumes/01-quantity/ch09-discipline-of-estimation/figures/fig-fermi-tree.tex}

\begin{archetype}[Scaling, canonical home]
The engineer's habit of placing every quantity at its order of magnitude
before reaching for the calculator. The habit has anchor points (a
person is $10^{2} \,\si{\kilogram}$; the Earth's radius is
$10^{7} \,\si{\meter}$; the speed of light is $10^{8.5} \,\si{\meter\per
\second}$; one year is $10^{7.5} \,\si{\second}$); it has decomposition
techniques (multiplicative factors, geometric subdivision, dimensional
balance); it has compounding rules (errors compound by multiplication
with the factor-of-three working target); and it has a closing move
(comparison to ground truth where ground truth exists). The archetype
arrived in chapter 1 of this volume and has been exercised in chapters
1, 2, 3, 4, and 5; this chapter codifies the habit and asks the reader
to perform fifty acts of it.
\end{archetype}

\section{Fermi problems}\pathtag{core}

A \engterm{Fermi problem} is a quantitative question answered by
decomposition before lookup. The reader may be able to find a published
value later; the discipline is to make and record the estimate first.
The genre is named for Enrico Fermi, who used decomposition both as a
teaching device and as a working tool. The habit predates Fermi; the
name has stuck because his standard, that an answer is worth defending
only if it comes with a method, has become the genre's standard.

The Fermi problem is the engineer's working unit of estimation. It has
a question, a decomposition, a small number of factors each estimated
from working knowledge, a multiplication, and a comparison to ground
truth where one exists. Its currency is the working engineer's mental
table of anchor points; its discipline is the habit of writing the
factors down before multiplying them.

\subsection{The shape of a Fermi solution}

Mahajan's \emph{The Art of Insight in Science and Engineering}
\cite{text:mahajan2014} is the canonical contemporary reference. He
organises Fermi-problem solutions around two structural moves:
divide-and-conquer (decompose the unknown into a product of factors
that the reader can each estimate) and abstraction (replace a
complicated function with a simpler one whose value the reader can
defend). We adopt the same structure and add a third move at the close:
postmortem (identify the factor that dominated the uncertainty, and
state what working knowledge would tighten the answer).

A worked Fermi solution then has five labelled parts.

\engterm{Question}. Stated explicitly, with the units of the answer
identified before any decomposition is attempted. A question whose
answer's units are unclear has not yet been stated as a quantitative
question.

\engterm{Decomposition}. The unknown $X$ is written as a product of
factors $X = \prod_{i} f_{i}$, with each factor a quantity the reader
can estimate from working knowledge. The decomposition is not unique;
different decompositions of the same question can produce estimates
with different dominant uncertainty sources.

\engterm{Factor estimates}. Each factor is estimated, with its
working assumption and its rough uncertainty range identified. A
factor estimated as $10$ with an admitted factor-of-two range is more
useful than a factor stated as $10$ without any range.

\engterm{Combination}. The factors are multiplied. Where the
combination is a sum or a more complex algebraic expression, the
algebra is done explicitly. The product carries a combined uncertainty
that, in the working approximation, multiplies the individual factor
uncertainties.

\engterm{Postmortem}. The estimate is compared to ground truth where
one is available. The factor that dominated the uncertainty is named.
The working knowledge that would tighten the estimate is identified.
The reader records the comparison, especially when the estimate was
wrong.

The five parts compose the answer. A Fermi solution that omits any one
of them is incomplete.

\subsection{Decomposition: divide-and-conquer}

Decomposition is the first hard move. The question is rarely
posed in a form that can be estimated directly; it has to be
rewritten as a product of factors that working knowledge can
each defend.

For \emph{how many cars in this city}, decompose as: city
population times cars per person.

For \emph{how much energy does the United States use per year},
decompose as: per-person power use times population times
seconds per year. With per-person continuous power $\sim
10\,\si{\kilo\watt}$ (the working anchor for an industrial
nation, sometimes called the "10 kW lifestyle"), $3.3 \times
10^{8}$ population, and $3.15 \times 10^{7}\,\si{\second}/
\text{yr}$, the answer is $10^{4} \times 3.3 \times 10^{8} \times
3.15 \times 10^{7} \approx 10^{20}\,\si{\joule}/\text{yr}$, or
$\sim 100\,\text{EJ}/\text{yr}$. The EIA's reported value
for $2023$ is $94\,\text{EJ}$. The estimate is correct to
within $\sim 10\%$ using nothing but three anchor points.

For \emph{how many seconds in a human lifetime}, decompose as:
$80\,\text{yr}$ times $365\,\text{day}/\text{yr}$ times
$86400\,\si{\second\per\day} \approx 2.5 \times 10^{9}\,\si{
\second}$. The answer is "a few gigaseconds", a useful working
anchor that the reader can use across many estimation problems.

For \emph{the mass of the Earth's atmosphere}, decompose as:
surface area $A_{\oplus} = 4\pi R^{2} \approx 5.1 \times 10^{14}
\,\si{\meter\squared}$ times the column mass per unit area
$= P/g \approx 10^{5}/10 = 10^{4}\,\si{\kilo\gram\per\meter
\squared}$, giving $5 \times 10^{18}\,\si{\kilo\gram}$. The
NIST-traceable value (atmospheric dry mass) is $5.15 \times
10^{18}\,\si{\kilo\gram}$. Three anchor points (Earth radius
$6378\,\si{\kilo\meter}$, atmospheric pressure $101\,\si{\kilo
\pascal}$, gravitational acceleration $9.81\,\si{\meter\per
\second\squared}$); the estimate is correct to one significant
figure.

For \emph{the number of grains of rice in one cubic metre},
decompose as: $1\,\si{\meter\cubed}$ divided by the volume of a
grain. The volume of a grain is decomposed further:
$5\,\si{\milli\meter} \times 2\,\si{\milli\meter} \times 2\,\si{
\milli\meter} \approx 2 \times 10^{-8}\,\si{\meter\cubed}$,
giving about $5 \times 10^{7}$ grains.

For \emph{the number of piano tuners in Chicago} (Fermi's
canonical problem, posed to physics graduate students at the
University of Chicago in the 1940s), decompose as: Chicago
population ($3 \times 10^{6}$) times pianos per household
(estimate $1$ in $30$ households has a piano) divided by piano
tunings per tuner per year (estimate $4 \times 200$ working days
$= 800$) times tunings per piano per year (estimate $1$). With
households per population ($1$ in $2.5$), the working count is
$3 \times 10^{6}/2.5/30 = 4 \times 10^{4}$ pianos, divided by
$800$ pianos per tuner-year, gives $50$ piano tuners. The
American Federation of Musicians' working count in 2023 for
Chicago metropolitan area: $\sim 60$ registered piano
technicians. The estimate is correct to a factor of $1.2$.

For \emph{the number of moles of water in the world's oceans},
decompose as: ocean volume $\approx 1.4 \times 10^{18}\,\si{
\meter\cubed}$ times density $10^{3}\,\si{\kilo\gram\per\meter
\cubed}$ divided by molar mass of water $18\,\si{\gram\per\mol}
= 1.8 \times 10^{-2}\,\si{\kilo\gram\per\mol}$, giving
$\sim 7.8 \times 10^{22}\,\si{\mol}$. The NIST value
($1.37 \times 10^{18}\,\si{\meter\cubed}$ ocean volume) gives
$\sim 7.6 \times 10^{22}\,\si{\mol}$, within $5\%$ of the
estimate. The order of magnitude is $\sim 10^{23}\,\si{\mol}$,
close to one Avogadro number per molecule. The anchor "the
ocean has about Avogadro's number of moles of water" is a
working aide-memoire.

The shape of a good decomposition is one in which each factor
sits between $10^{-2}$ and $10^{2}$ of working knowledge anchor
points. A factor that requires the reader to estimate a quantity
orders of magnitude outside their experience is a sign that the
decomposition needs another step. The five working examples
above each used between three and five anchor points; each
anchor point was a quantity the practising engineer should
internalise (population by country, atmospheric pressure,
typical household composition, ocean volume). The chapter's
final section lists fifty problems where the reader practices
the working moves on quantities ranging from physical to urban
to biological.

\subsection{Abstraction: replace the complicated with the simple}

Abstraction is the second hard move. The question may involve a
function the reader cannot evaluate, a population the reader does not
know, a relation the reader cannot integrate (Volume II takes up
integration; this chapter does not). Abstraction replaces the
complicated object with a simpler one that captures the dominant scale.

A non-uniform population is replaced by its mean times its count. A
shaped object is replaced by a box of equivalent dimensions. A
distribution is replaced by its peak. A time-varying signal is
replaced by its time average. A nonlinear function near a working
point is replaced by its tangent line. Each replacement loses
information; the discipline is to choose the replacement so that the
information lost is not the information the question depends on.

The habit of abstraction is what makes Fermi estimation possible
without calculus. The reader who wants to estimate the volume swept by
a fan blade does not integrate the swept area over the rotation; the
reader replaces the blade by a flat disk of the same outer radius and
multiplies by the rotation count. The abstraction discards the
turbulent details, which are not what the volume question depends on.

\subsection{Combination and uncertainty propagation}

A product of factors compounds its uncertainties in log space. If
$X = \prod_{i} f_{i}$, then $\ln X = \sum_{i} \ln f_{i}$. For
independent factor uncertainties, the log-uncertainty obeys
\[
\sigma_{\ln X}^{2} = \sum_{i} \sigma_{\ln f_{i}}^{2}.
\]
A factor-of-$1.5$ uncertainty corresponds to $\sigma_{\ln f} \approx
\ln(1.5) \approx 0.41$ if we treat the factor as a one-sigma log-width.
Four such factors combine to $\sigma_{\ln X} \approx 0.82$, a
multiplicative one-sigma factor of $e^{0.82} \approx 2.3$. Seven such
factors combine to $e^{\sqrt{7}\,\ln(1.5)} \approx 2.9$. The
approximation is only a model of independent factor errors, but it
teaches the right habit: a long decomposition widens the result.

The implication is structural. A Fermi decomposition with three or
four factors typically produces a result good to within a factor of
three. A decomposition with seven factors produces a result good only
to within an order of magnitude, even if every factor is well
estimated. The reader who is tempted to break a question into many
small factors should check that the multiplicative compounding does
not exceed the precision the question requires.
\Cref{fig:vol01:ch09:log-uncertainty} plots the combined factor
uncertainty against the number of factors for two per-factor working
widths and reads as the design chart for how deep a decomposition the
chapter's working target can support.

\input{volumes/01-quantity/ch09-discipline-of-estimation/figures/fig-log-uncertainty-stacking.tex}

The Monte Carlo verification in
\repopath{volumes/01-quantity/ch09-discipline-of-estimation/code/monte_carlo_fermi.py}
samples random factor errors and confirms the predicted log-quadrature
combined width to within a few percent for $k$ up to seven; the
script also prints the combined one-sigma factor for each $k$, which
is the number a reader uses to translate a stated decomposition into
a working uncertainty band.

\paragraph{Worked example: world primary energy.}
World annual primary energy consumption decomposes as world population
times per-capita primary power times seconds per year, three factors
with per-factor uncertainty around $1.3\times$ (population is known to
about $1.05\times$; per-capita power varies regionally by a factor of
five but the world-average value is known to about $1.2\times$;
seconds per year is exact). The combined log-quadrature uncertainty is
$\sigma_{\ln X} = \sqrt{3} \times \ln 1.3 \approx 0.45$, a
multiplicative one-sigma factor of $e^{0.45} \approx 1.57$. The
nominal estimate, $8 \times 10^{9} \times 2500 \,\si{\watt} \times
3.15 \times 10^{7} \,\si{\second} \approx 6 \times 10^{20}
\,\si{\joule}$, lies within a factor of two of the Energy Institute's
$6.0 \times 10^{20} \,\si{\joule}$ figure for 2022
\cite{data:ch09-bp-statistical-2024}, well inside the predicted band.

\paragraph{Worked example: a seven-factor estimate that loses
its grip.} The annual mass of CO$_2$ released by all
commercial-aviation passenger flights decomposes as: world
population $\times$ fraction who fly per year $\times$ flights
per flying person per year $\times$ average passenger-km per
flight $\times$ fuel burn per passenger-km $\times$ CO$_2$ per
unit fuel burned $\times$ correction factor for short-haul vs
long-haul mix. Seven factors. Per-factor uncertainties: world
population $1.02\times$ (well known), fraction who fly $1.5\times$
(rich-world $\sim 50\%$, world-average $\sim 11\%$, varies
strongly by year and income band), flights per flying person
$1.5\times$, average passenger-km $1.3\times$, fuel burn per
passenger-km $1.3\times$ (modern jets $\sim 30\,\text{g CO}_2$
per passenger-km, regional jets up to $90$, fully loaded vs
half-loaded matter), CO$_2$ per kg jet fuel ($3.16$ exact,
$1.0\times$), short/long correction $1.3\times$. Log-quadrature:
$\sigma_{\ln X} = \sqrt{2 \times 0.02^{2} + 4 \times 0.41^{2} +
0.27^{2}} \approx 0.92$, a multiplicative one-sigma factor of
$e^{0.92} \approx 2.5$. The estimate $8 \times 10^{9} \times
0.11 \times 1.5 \times 2500\,\text{km} \times 30\,\text{g CO}_2/
\text{(pax km)} \approx 1.0\,\text{Gt CO}_2$ straddles the IEA's
reported $0.96\,\text{Gt}$ for 2023, but the $\pm 2.5\times$
band runs from $\sim 0.4$ to $\sim 2.5\,\text{Gt}$. The reader
who needs $\pm 30\%$ precision on aviation emissions must
collapse the decomposition; the practical move is to anchor on
a single industry-aggregated figure (ICAO's reported tonne-km
per year) and convert, sacrificing first-principles
decomposability for one-anchor precision.

\subsection{Postmortem: which factor dominated}

The postmortem is the move that turns the estimate into a learning
event. Once the answer is in hand and ground truth, where available,
has been compared, the reader identifies the factor whose uncertainty
dominated the result. That factor is the one whose anchor point in the
reader's working knowledge needs improvement. The reader records the
factor and, where possible, the truer value; the next time a similar
question arises, the reader's estimate starts from a tighter anchor.

The Chicago piano-tuner estimate's dominant uncertainty was the
fraction of households with a piano. A reader who learns, from this
estimate, that the fraction is closer to $1/40$ than to $1/4$ has
acquired an anchor that will sharpen any subsequent estimate involving
musical-instrument ownership. The Fermi practice is a directed
learning machine; the postmortem is what directs the learning.

The postmortem reads best as a budget rather than as a single named
culprit. Each factor in a decomposition carries a share of the total
log-variance $\sigma_{\ln X}^{2}$, and the shares sum to one.
\Cref{fig:vol01:ch09:variance-budget} draws the budget for the
piano-tuner estimate: the households-with-piano factor carries almost
half of it, the tunings-per-piano factor carries most of the rest, and
the population factor, though it fixes the answer's magnitude, carries
almost none of the uncertainty because it is known to a few percent.
The working rule the budget makes explicit is that effort spent
narrowing a factor should be proportional to that factor's share, not
spread evenly. A reader who spends an afternoon refining the population
figure has improved the estimate by less than a percent; the same
afternoon spent finding a survey of piano ownership rates would cut the
dominant share and tighten the whole result.

\input{volumes/01-quantity/ch09-discipline-of-estimation/figures/fig-variance-budget.tex}

\subsection{Three more decompositions, worked to the postmortem}

The four moves are learned by repetition on quantities the reader can
each defend from working knowledge. Three further worked examples
extend the range of domains and each closes on the postmortem, so the
reader sees the dominant-factor identification performed rather than
only described.

\paragraph{Worked example: rain falling on a city in a storm.}
\emph{How much water, in tonnes, falls on a metropolitan area of
$1{,}200 \,\si{\kilo\meter\squared}$ during a heavy one-hour
thunderstorm?} The decomposition is area times rainfall depth times
water density. A heavy thunderstorm deposits on the order of $20 \,
\si{\milli\meter}$ of rain in an hour ($2 \times 10^{-2} \,\si{\meter}$,
between the $10 \,\si{\milli\meter}$ of a soaking shower and the $50 \,
\si{\milli\meter}$ of a flash-flood cloudburst). The volume is
$1.2 \times 10^{9} \,\si{\meter\squared} \times 2 \times 10^{-2}
\,\si{\meter} = 2.4 \times 10^{7} \,\si{\meter\cubed}$, and at
$1{,}000 \,\si{\kilo\gram\per\meter\cubed}$ the mass is
$2.4 \times 10^{10} \,\si{\kilo\gram} = 2.4 \times 10^{7}
\,\si{\tonne}$. The unit check anchors the scale: twenty-four million
tonnes of water is roughly the displacement of two hundred large
container ships, delivered onto the city in one hour, which is why
urban drainage is a serious civil-engineering load and not an
afterthought. The postmortem names the rainfall-depth factor as the
dominant one; it varies by a factor of five between a shower and a
cloudburst, while the area is known exactly and the density is fixed.
An engineer sizing a storm drain who carries this estimate reads a
published design-storm intensity against an independent expectation
and catches a units slip (millimetres per hour read as centimetres per
hour) in seconds.

\paragraph{Worked example: heartbeats in a lifetime, across species.}
\emph{How many times does a human heart beat in a lifetime, and how
does the count compare with a mouse and an elephant?} The human
decomposition is resting rate times seconds per year times lifespan:
$70 \,\text{min}^{-1} = 1.17 \,\si{\per\second}$, times $3.15 \times
10^{7} \,\si{\second}/\text{yr}$, times $75 \,\text{yr}$, gives
$\approx 2.8 \times 10^{9}$ beats. A mouse beats at $\approx 600 \,
\text{min}^{-1}$ and lives $\approx 2 \,\text{yr}$: $10 \,\si{\per
\second} \times 3.15 \times 10^{7} \times 2 \approx 6 \times 10^{8}$
beats. An elephant beats at $\approx 30 \,\text{min}^{-1}$ and lives
$\approx 65 \,\text{yr}$: $0.5 \,\si{\per\second} \times 3.15 \times
10^{7} \times 65 \approx 1.0 \times 10^{9}$ beats. The three counts
sit within a factor of five of one another despite a lifespan range of
thirty and a heart-rate range of twenty, because the two factors move
in opposite directions and their product is stable. The postmortem's
lesson is a structural one: when a decomposition's factors are
negatively correlated across cases, the product is far tighter than
either factor, and the estimate that looks fragile in its inputs is
robust in its output. The human count runs high against the mammalian
$\approx 10^{9}$ working anchor, a signal that human lifespan has been
extended by engineering (sanitation, medicine, nutrition) well past
the mammalian allometric line.

\paragraph{Worked example: the reciprocal check on a per-unit figure.}
\emph{A vendor quotes a data-centre power draw of $30 \,\si{\mega\watt}$
and claims it serves $500{,}000$ concurrent users; is the per-user
power plausible?} The estimate here runs backward from a claimed total
to a per-unit figure, which is the fastest form of sanity check on a
quoted aggregate. Dividing gives $3 \times 10^{7} \,\si{\watt} / 5
\times 10^{5} = 60 \,\si{\watt}$ per concurrent user. An independent
expectation from working anchors puts the datacentre-side power of a
single streaming or conferencing session at a fraction of a watt to a
few watts, not tens of watts; $60 \,\si{\watt}$ per user is one to two
orders of magnitude high. The postmortem locates the discrepancy in
the definition of ``concurrent user'': the $30 \,\si{\mega\watt}$ is
the facility's provisioned peak, and the $500{,}000$ is a
simultaneously-connected count, but the provisioned peak also covers
cooling overhead, redundant capacity, and storage that is not
attributable per active session. The reciprocal check does not prove
the vendor wrong; it flags a per-unit figure that disagrees with the
working anchor by more than the factor-of-three target and forces the
definitions to be pinned before the number is used in a decision. The
reciprocal check is the cheapest sanity move in the chapter: any quoted
aggregate divided by its quoted count yields a per-unit figure that a
working anchor can test in one line.

\subsection{An applied vignette: the reactor coolant back-of-envelope}

A working instance ties the four moves to a decision an engineer makes
under time pressure. During a design review a junior engineer presents
a calculated cooling-water flow rate for a proposed $50 \,\si{\mega
\watt}$ industrial process heater: $12 \,\si{\liter\per\second}$. The
reviewer, who has not seen the calculation, runs an independent
estimate in the margin before the presenter reaches the next slide.
The heat to be removed is $50 \,\si{\mega\watt}$. Water carries
$c_{p} = 4{,}184 \,\si{\joule\per\kilogram\per\kelvin}$, so a $10 \,
\si{\kelvin}$ temperature rise across the heat exchanger removes
$4{,}184 \times 10 \approx 4.2 \times 10^{4} \,\si{\joule}$ per
kilogram of water. The required mass flow is $5 \times 10^{7} \,\si
{\watt} / 4.2 \times 10^{4} \,\si{\joule\per\kilogram} \approx
1{,}200 \,\si{\kilogram\per\second}$, which at water density is
$\approx 1{,}200 \,\si{\liter\per\second}$. The reviewer's estimate
and the presented number disagree by a factor of one hundred.

The disagreement is a release-blocking signal, and the four moves
locate it. The question was stated with the units fixed (litres per
second of coolant). The estimate was recorded before the presented
number was interrogated, which preserves it as a blind comparison. The
comparison is quantitative: a factor of one hundred is far outside the
factor-of-three target. The postmortem, run against the presenter's
worksheet, finds a unit-conversion error: the presenter had computed
the flow for a $1{,}000 \,\si{\kelvin}$ temperature rise, having
entered the heater's outlet gas temperature where the water-side
temperature rise belonged. The estimate did not need to be precise to
be decisive; a margin calculation good to a factor of two flagged a
calculation wrong by a factor of one hundred, and the design review
held the number until the worksheet was corrected. The vignette is the
chapter's discipline compressed into the five minutes a reviewer has
during a live presentation, and it is the setting in which the habit
pays for itself most visibly.

\begin{mastery}
The reciprocal check and the two-route check are the same move seen
from two directions, and both are special cases of a
\emph{dimensional consistency of magnitudes} argument. A quoted total
$X$ with a quoted count $N$ yields a per-unit figure $X/N$ that a
working anchor tests; a quantity computed by route A and by route B
yields a ratio $X_{A}/X_{B}$ that unity tests. In each case the test
statistic is a dimensionless ratio, and the discipline is to require
the ratio to sit within the factor-of-three target before the number
is accepted. The generalisation the reader carries into Volume II is
that every sanity check in the chapter can be written as a
dimensionless ratio compared against an expected value: order of
magnitude against the recorded estimate, per-unit against the working
anchor, route A against route B, limiting case against the closed
form. The reader who trains the habit of forming the ratio, rather
than comparing two raw numbers, gains a check that is invariant to the
units the two numbers happen to be reported in, which is exactly the
invariance a units slip destroys. Volume II Chapter 2 formalises the
same move as the logarithmic distance $|\ln(X_{A}/X_{B})|$, whose
being small is the precise statement of ``the two routes agree.''
\end{mastery}

\section{Estimation before calculation: a habit}\pathtag{core}

The discipline that makes estimation useful is the order in which it is
applied. The estimate is recorded before the calculation begins. Once
the calculation has produced a number, that number anchors judgement;
later factors tend to drift toward it. The estimate-before-calculate
discipline defends the calculation against silent error by preserving
a blind comparison value.

\subsection{Why the order matters}

Confirmation bias is the human tendency to interpret new evidence as
confirming existing beliefs. The engineer who has just calculated a
result is in possession of a strong existing belief about the answer.
An estimate produced after the calculation is filtered through that
belief: factors that would have been chosen as $0.5$ are chosen as
$0.4$ to bring the estimate closer to the calculated number; factors
that would have been chosen as $1.5$ are chosen as $1.7$. The estimate
becomes a story that the calculation has commissioned.

The defence is procedural. The estimate is performed first, written
down, and dated. The calculation follows. The comparison is made
against the recorded estimate. When the two disagree by more than the
estimate's stated range, the reader investigates the calculation
before adjusting the estimate.

The same procedural defence appears in many places in engineering: the
control sample taken before the experiment, the design review held
before the prototype is fabricated, the load case stated before the
analysis is run. In each case, a record made before a result is the
discipline that protects against the human tendency to align records
with results retroactively.

\subsection{The estimate as a sanity check}

A calculated number that disagrees with the estimate by more than a
factor of three is, by default, suspect. Three failure modes account
for most such disagreements.

\engterm{Unit error}. The calculation has multiplied or divided by a
unit conversion factor in the wrong direction. This is the most
frequent cause of factor-of-thousand and factor-of-million
discrepancies between estimate and calculation. A spreadsheet that
expected millimetres but received metres has the wrong answer by a
factor of $1000$; one that expected litres but received cubic metres
has the wrong answer by a factor of $1000$. Volume I Chapter 1 is the
canonical home of the unit-error case.

\engterm{Sign error}. A subtraction that was meant to be an addition;
a logarithm whose argument was inverted; an exponent whose sign was
flipped. The estimate produces a positive result of order $10^{4}$;
the calculation produces a negative result of order $10^{4}$. The
factor-of-magnitude check passes; the sign check fails. The reader
who relies only on order-of-magnitude comparison will miss this
class of error.

\engterm{Decomposition error}. The estimate's decomposition was wrong:
the reader missed a factor, or double-counted a factor, or used the
wrong functional form. This is the hardest disagreement to resolve,
because the disagreement says only that one of the two arguments is
wrong, and the engineer must investigate both. The investigation often
reveals that the calculation is right and the estimate's decomposition
was not what the question called for. The reader corrects their
decomposition and learns something about the question.

The first two classes are caught by the estimate within minutes. The
third class can take hours. The estimate-before-calculate discipline
makes all three visible; without it, the unit error and the sign error
slip silently into the report.

\subsection{The habit's place in a working session}

A working engineering session that uses the discipline has the
following shape. The question is stated, with the units of the answer
identified. The estimate is performed and recorded, with its
decomposition and its expected range. The calculation is performed.
The two are compared. Where they agree within the estimate's range,
the calculation is taken as the working answer; where they disagree,
the investigation begins.

The habit costs about ten minutes per estimate in the first weeks of
practice. After several months it often costs two minutes; after
enough repetition it becomes part of how the engineer reads the
question. The cost is small enough that the check belongs before any
calculation whose result will leave the desk.

\section{Sanity-checking calculated results}\pathtag{core}

Once the calculation is done, the estimate is one of several
sanity checks the engineer applies to the result before passing it
on. The other checks complement the estimate; together they catch
errors that any single check would miss.

\subsection{The order-of-magnitude check}

The first check, performed against the recorded estimate, is the one
described in the previous section. The calculated result and the
estimate should agree within the estimate's stated range. When they
do not, the engineer investigates before reporting.

\subsection{The dimensional check}

Volume I Chapter 2 introduced dimensional analysis as the syntax of
quantitative reasoning. The dimensional check applied to a calculated
result is the syntactic counterpart to the order-of-magnitude check's
semantic role. A result that has the right magnitude but the wrong
dimensions is, with rare exceptions, wrong. A result that has the
wrong magnitude but the right dimensions is more often wrong than
right.

The dimensional check is performed last, on the final algebraic
expression, before any numerical substitution. The expression's
dimensions are computed from the dimensions of its inputs and
compared to the expected dimensions of the answer. When they disagree,
the calculation is wrong before any number has been computed.

\subsection{The sign check}

Most engineering quantities have a definite sign. A mass is positive.
A length is positive. An energy expended is positive. A heat flow
into a system is positive when the convention is that heat flows in;
negative when the convention is that heat flows out. A torque is
signed; a force component along a coordinate axis is signed.

The sign check is done by going through each term in the result and
asking whether its sign matches the physical situation. A force
calculated to point upward when gravity should pull it down has
either a sign error or a coordinate-system mismatch; the calculation
must be re-examined. A heat that flows from cold to hot violates the
second law; the calculation is wrong.

The sign check is fast. It catches a class of errors that the
order-of-magnitude check misses and that the dimensional check often
misses.

\subsection{The limiting-case check}

A calculation that produces a defensible answer for general inputs
should produce a defensible answer for limiting-case inputs. Send one
input to zero; send another to infinity; set two inputs equal; impose
a symmetry. The result should reduce to a defensible limiting form.

The pendulum period $T = 2\pi \sqrt{L/g}$ has the limiting forms:
$T \to 0$ as $L \to 0$ (a pendulum with no length has no period);
$T \to \infty$ as $g \to 0$ (a pendulum with no gravity does not
oscillate); and $T = 2\pi \,\si{\second}$ when $L/g = 1 \,\si{\second
\squared}$. Each limit can be checked against intuition and dimensions.
The
calculation that has the wrong functional form often passes the
order-of-magnitude check at one input and fails it at another; the
limiting-case check exposes the discrepancy.

\subsection{The independent-method check}

The most powerful sanity check is to compute the same quantity by an
independent method. Volume I Chapter 4's project asked the reader to
measure a building's height by three independent methods. The
discipline applies equally to calculations: a result computed by one
algebraic route can be checked by computing the same result by a
different route. When the two routes agree, the result has been
checked against arithmetic error; when they disagree, the engineer
investigates.

The independent-method check is the most labour-intensive of the
sanity checks because it requires building two paths to the same
result. For safety-critical work, the second path is normally
performed by a different engineer. Shared inputs, such as
requirements, material properties, and environmental assumptions,
are named explicitly and audited because common inputs can create
common-mode error.

\subsection{The reasonable-result check}

The final check is the engineer's judgement of whether the result is
reasonable for the system being modelled. A calculated bridge
deflection of $0.001 \,\si{\milli\meter}$ under design load is
suspect not because the dimensional check fails but because the
working engineer knows that a real bridge under design load deflects
by several centimetres. A calculated motor current of
$10{,}000 \,\si{\ampere}$ for a household appliance is suspect
because the household electrical supply does not deliver currents at
that scale. The reasonable-result check uses the engineer's working
knowledge of the system to spot a result that the more abstract
checks have all passed but that working experience says cannot be
right.

The check is the slowest to develop, because it requires accumulated
working experience that no textbook can supply. The book can describe
the check; only practice on real systems can build the experience the
check uses.

\subsection{Three worked sanity checks: cases where the estimate caught the
error}

The sanity-check sequence is most concrete when followed through a
case in which a working engineer caught an error before fabrication,
licensing, or publication. Three cases that span experimental
physics, nuclear licensing, and data-analytics software make the
sequence's payoff visible.

\paragraph{Case 1: a particle-physics jet-energy scale check
(CMS, $2018$).} The Compact Muon Solenoid (CMS) detector at the
Large Hadron Collider measures jet energies by summing
calorimeter cell deposits. The detector's jet-energy scale (JES)
calibration is the dominant systematic in many CMS analyses; the
$2018$ update to the calibration carried a $\approx 1\%$
uncertainty at high transverse momentum
\cite{paper:ch09t3b-cms-jes-2017}. An analysis-stage Fermi check
on the calibrated jet energy proceeds: a proton-proton collision
at $\sqrt{s} = 13 \,\si{\tera\electronvolt}$ deposits up to
$\sim 10^{4} \,\si{\giga\electronvolt}$ of energy into the
detector when both partons carry a substantial fraction of the
beam momentum, but a representative high-$p_{T}$ dijet event in
the working analysis sits at $p_{T} \sim 500$--$1{,}500 \,
\si{\giga\electronvolt}$. The hadronic calorimeter (HCAL) covers
$|\eta| < 3$ with cells of solid angle $\Delta \eta \Delta \phi
\approx 0.087 \times 0.087$; the deposited energy density in a
jet cone of radius $R = 0.4$ is therefore
$E_{T}^{\text{cell}} / E_{T}^{\text{jet}} \sim 1/N_{\text{cells}}$
with $N_{\text{cells}} \approx \pi (0.4/0.087)^{2} \approx 70$.
An engineer who has run this Fermi check on the first plotted
calibrated jet of an analysis catches a JES misconfiguration that
multiplies the displayed jet energy by the wrong inverse
correction (the JES correction $f_{\text{JES}} \approx 1.05$ at
the working $p_{T}$, but applied as $1/f_{\text{JES}}$ produces a
mis-scaled jet by a factor of $\approx 1.10$, large enough to
shift a published cross-section by twice the analysis's stated
systematic). The CMS collaboration's internal review process
includes exactly such a Fermi check at the analysis-note stage;
the reviewer's working number is the published JES correction
factor at the analysis $p_{T}$, and the displayed corrected jet
energy is required to agree within the calibration's systematic
\cite{paper:ch09t3b-cms-jes-2017}. The case is the discipline at
the highest-precision end of experimental physics; the same
sanity-check sequence applies, with the published reference
numbers as the calibration anchor.

\paragraph{Case 2: a nuclear-reactor containment-pressure check
(General Electric Mark I/II/III, early $1970$s).} The General
Electric BWR Mark I containment design used a pressure-suppression
pool to absorb steam released by a loss-of-coolant accident
(LOCA), with the containment volume sized to keep the peak
post-accident pressure below the design pressure of $62 \,
\text{psig} = 0.427 \,\si{\mega\pascal}$ above ambient
\cite{paper:ch09t3b-nureg-cr-7222-2016}. A Fermi estimate of the
peak pressure proceeds from a conservation argument: a reactor
operating at $P_{\text{th}} = 3{,}000 \,\si{\mega\watt}_{\text{th}}$
releases $\Delta E_{\text{LOCA}} \approx M_{\text{coolant}} \cdot
c_{p} \cdot (T_{\text{op}} - T_{\text{sat,atm}})
+ M_{\text{coolant}} \cdot h_{fg}$ on rapid depressurisation;
with $M_{\text{coolant}} \approx 2 \times 10^{5} \,
\si{\kilogram}$ and $h_{fg} \approx 2.26 \,
\si{\mega\joule\per\kilogram}$, the total enthalpy release is
$\sim 5 \times 10^{11} \,\si{\joule}$. Absorbed into a
suppression-pool water mass of $\sim 3 \times 10^{6} \,
\si{\kilogram}$ at $c_{p,\text{water}} = 4{,}184 \,
\si{\joule\per\kilogram\per\kelvin}$, the pool temperature rise is
$\Delta T \approx 5 \times 10^{11} / (3 \times 10^{6} \times
4{,}184) \approx 40 \,\si{\kelvin}$, and the resulting equilibrium
pressure is dominated by the dry-well gas-phase compression
during blow-down. The dry-well to wet-well volume ratio sets the
peak pressure to within a factor of two of design; an estimate
that disagrees by more than a factor of two flags a geometric
input error. Documented Mark I design-history records show the
working containment community caught and corrected such estimates
during the $1972$ review cycle; the eventual NUREG-0978 record of
the post-$1979$ containment-loads programme codifies the Fermi
sequence as a required containment-margin check
\cite{paper:ch09t3b-nureg-cr-7222-2016}. The estimate-before-
calculation sequence at the licensing-document stage caught
geometric and thermodynamic input errors before the detailed
finite-element pressure-transient code was run, in exactly the
discipline this chapter has named.

\paragraph{Case 3: a SQL query-result Fermi check (working
software engineering, $2010$s onward).} A working data-analysis
session opens with a SQL query against an operational warehouse;
the query returns a single number, and a derived report
propagates the number to a decision. The Fermi check before the
query is run is: from the working anchors and the question's
context, what should the number be? An estimate of the number of
US-resident customers a SaaS company has, with a known $\$10$M
annual recurring revenue and an average revenue per user of
$\$50$/month, is $10^{7} / (50 \times 12) \approx 1.7 \times
10^{4}$ active customers. A SQL query
\verb|SELECT COUNT(*) FROM customers WHERE country = 'US'| that
returns $1.7 \times 10^{6}$ is wrong by a factor of $100$. The
most frequent root causes are documented in the data-engineering
literature \cite{text:ch09t3b-kimball-2013}:

\begin{itemize}\setlength\itemsep{2pt}
\item A missing \verb|WHERE status = 'active'| clause includes
churned and trial accounts.
\item A \verb|JOIN| against an event-log table without
\verb|SELECT DISTINCT| inflates the row count by the average
number of events per customer.
\item A \verb|country IN ('US','USA','United States')|
mishandling treats a single customer with multiple normalised
country values as multiple customers.
\item A cohort-window \verb|WHERE created_at < now()| includes
deleted records that should have been excluded by a
\verb|deleted_at IS NULL| predicate.
\end{itemize}

The engineer who has not performed the Fermi check before running
the query has no working defence against any of these errors;
the query's result is reported, and the decision that depends on
it is taken from a number that was wrong by a factor of $10$ or
$100$. The engineer who has performed the Fermi check catches the
discrepancy in seconds, investigates the query, and corrects it
before the report is distributed. The data-engineering literature
records this as the canonical query-review discipline
\cite{text:ch09t3b-kimball-2013, web:ch09t3b-wickham-tidy-2014}:
a query whose result disagrees with the order-of-magnitude
estimate by more than a factor of three is treated as wrong by
default, and the joins, filter predicates, and table semantics
are audited before the result is accepted. The working ratio of
caught-to-uncaught errors in teams that run the Fermi check is
high enough that several large data-platform vendors document
the discipline as a step in their published query-review style
guides.

The three cases span six orders of magnitude in the underlying
quantity, three working domains (experimental physics, nuclear
engineering, data engineering), and three decades of working
practice. The sanity-check sequence is the same in each: estimate
first, compute second, compare, and investigate any disagreement
larger than the estimate's stated range before the result leaves
the desk.

\section{Estimation as a design habit}\pathtag{standard}

The chapter so far has placed estimation at three points: before a
calculation, against a calculation, and as a sanity check on a
calculation's result. A wider use is the one engineers acquire after
several projects: the early-design estimate that decides whether a
proposed system is worth analysing in detail. The design habit
embeds estimation in the project lifecycle, with three distinct
applications and three different stopping rules.

\subsection{When the estimate enters the project}

A design project passes through phases at which the available
information about the system changes by orders of magnitude. The
estimation discipline attaches differently at each phase, and the
working engineer who places the estimate at the wrong phase loses
most of its benefit.

\engterm{Pre-feasibility}. The proposed system exists as a
specification and a few key parameters. The estimate at this phase
asks whether the proposed system is physically possible. A
proposed battery whose stored energy exceeds the chemical energy
of its mass times the most energetic available redox couple is
not feasible regardless of clever engineering; the pre-feasibility
estimate rules it out in a single line. The estimate also asks
whether the proposed system's cost, power, mass, or volume is
within an order of magnitude of what comparable systems achieve;
when it is not, the design intent has either an undisclosed
breakthrough assumption or an undisclosed cost-of-failure. The
estimate's stopping rule at pre-feasibility is impossibility: a
proposed system that the estimate identifies as infeasible by a
factor of three is not taken to the next phase without an
explanation of which factor in the estimate is wrong.

\engterm{Conceptual design}. The system has a partitioned
architecture and budget allocations to each part. The estimate at
this phase checks consistency between the partitioned budgets and
the system-level requirement. A spacecraft power budget that
allocates more power to housekeeping than the solar array can
generate has an inconsistency the estimate exposes before any
detailed engineering begins. The stopping rule at conceptual
design is consistency: parts must sum to a system that the
top-down estimate identifies as feasible at the system level.

\engterm{Detailed design}. The system has a working specification
and parts that are being engineered to it. The estimate at this
phase is the sanity check the chapter's earlier sections trained.
Calculated values are checked against the estimate; large
discrepancies are investigated before the design proceeds. The
stopping rule at detailed design is the factor-of-three
disagreement: any calculated value that disagrees with the
estimate by more than a factor of three holds the design until
the disagreement is resolved.

The placement of the estimate at the project's earliest phase is
the one that saves the most. A pre-feasibility estimate costs
hours; a conceptual-design estimate costs days; a detailed-design
estimate costs weeks. The same disagreement caught at
pre-feasibility versus at detailed design is two to three orders
of magnitude apart in cost-of-rework. The discipline's economic
case is concentrated at the early phases.

\subsection{The estimate's three uses}

Inside a single design project the same estimate serves three
distinct uses. The reader who treats them as the same use loses
the discipline's specificity.

\engterm{Sanity check}. The estimate is performed against a
calculated, measured, or simulated quantity, and the comparison
flags disagreement. The use is what sections~9.3 and~9.4 have
trained. The action on a flag is investigation of the
disagreement; the estimate by itself is not the design decision.

\engterm{Risk flag}. The estimate is performed against a budget,
specification, or requirement, and the comparison flags whether
the budget is feasible. The use is the conceptual-design
consistency check. The action on a flag is reallocation of the
budget or reconsideration of the requirement; the estimate
identifies a risk that the requirement may not be met without
budget renegotiation. The risk flag is the move by which an
estimate transitions from technical practice to engineering
management.

\engterm{Stopping rule}. The estimate is performed against the
existence of a proposed system, and the comparison decides
whether the proposed system is worth further engineering effort.
The use is the pre-feasibility check. The action on a flag is
to stop the project, or to redirect it toward a different
architecture. The stopping rule is the most consequential of the
three uses because it determines whether engineering effort is
allocated to the project at all.

The three uses share the chapter's four moves (question,
estimate, comparison, postmortem) but differ in what they
compare against and in what action a flag triggers. The reader
who is producing an estimate names which use the estimate is
serving; the reader receiving an estimate asks the same
question.

\subsection{Worked case: the Boeing 787 battery thermal-runaway energy}

The case is chosen because the estimate, performed as a
pre-feasibility check on the battery's containment requirement,
would have flagged the design gap that the in-service incidents
later exposed. The reconstructed estimate uses information
available to Boeing and to the FAA before the 2007 special
condition was granted; the reconstruction is hypothetical at the
level of who would have performed the estimate, but the inputs
and the conclusion are entirely defensible from the available
literature.

\paragraph{Step 1: the question.}
A single cell of the 787's auxiliary-power-unit lithium-ion
battery is a $32 \,\text{V}/8$-cell-series, $75 \,\si{\ampere\hour}$
pack manufactured by GS Yuasa. The per-cell capacity is
$75 \,\si{\ampere\hour}$ at $3.7 \,\si{\volt}$ nominal. The
question for the pre-feasibility estimate: how much energy can
one cell release into the surrounding airframe in the worst
plausible failure mode, in joules, and is the containment
designed for that energy?

\paragraph{Step 2: the bare Fermi estimate.}
The electrochemical energy stored in one cell is
\[
E_{\text{stored}} = 75 \,\si{\ampere\hour} \times 3.7 \,\si{\volt}
\times 3600 \,\si{\second\per\hour}
\approx 1.0 \,\si{\mega\joule}.
\]
A familiar unit comparison: $1 \,\si{\mega\joule} \approx 240
\,\text{kcal} \approx 0.24$~food calorie thousands. The hand
grenade analogy that turned up in the post-incident press is
quantitatively defensible at the order-of-magnitude scale: an
M67 fragmentation grenade contains approximately $180 \,\text{g}$
of Composition~B explosive at a working detonation energy of
about $5 \,\text{kJ/g}$, or $\approx 0.9 \,\si{\mega\joule}$
\cite{paper:ch09t3-williard-pecht-2013}. The stored-energy
comparison is order-of-magnitude consistent.

\paragraph{Step 3: the worst-plausible-failure correction.}
The bare stored-energy estimate is a lower bound on the
energy a cell can release in thermal runaway, because the
runaway event also releases the cathode oxygen and the
electrolyte hydrocarbon fuel through exothermic decomposition.
The post-incident calorimetric measurements on identical
LVP65 cells reported by the JTSB and by independent academic
investigators give a total released energy of approximately
$7 \,\si{\mega\joule}$ to $8 \,\si{\mega\joule}$ per cell in
the worst-case venting mode, dominated by the combustion of
released electrolyte and cathode oxygen
\cite{acc:ntsb-aar14-01-787-battery,paper:ch09t3-williard-pecht-2013}.
The worst-plausible-failure correction is therefore a factor of
$\sim 7$ over the bare stored-energy estimate.

\paragraph{Step 4: the pre-feasibility comparison.}
The 2007 FAA Special Condition 25-359-SC required the
battery installation to demonstrate single-cell thermal-runaway
containment but did not require demonstration that the eight-
cell enclosure could contain a multi-cell propagated event
\cite{acc:ntsb-aar14-01-787-battery}. The single-cell
energy of $7 \,\si{\mega\joule}$, multiplied across the
plausible propagation count (the post-event analysis showed
cells $5$, $6$, $3$, and $7$ entered runaway in succession,
giving four cells in the worst observed event), reaches
$\approx 30 \,\si{\mega\joule}$. The containment-box mass and
its specific heat give the temperature rise the box would
experience under that energy release; the box was a thin-walled
aluminium enclosure whose mass and heat capacity rule out
absorbing $30 \,\si{\mega\joule}$ without venting hot gases
into the electronic-equipment bay
\cite{acc:ntsb-aar14-01-787-battery}.

\paragraph{Step 5: the design action the estimate would have
triggered.}
A pre-feasibility estimate that compares the worst-plausible-
failure energy of the eight-cell pack ($\approx 30
\,\si{\mega\joule}$) against the containment box's absorptive
capacity ($\lesssim 1 \,\si{\mega\joule}$ for the as-designed
box) identifies a design gap of more than an order of
magnitude. The action on the flag is either to redesign the
containment to absorb the worst-plausible-failure energy or to
restrict the failure mode to single-cell events (the path the
2007 special condition implicitly assumed, but did not verify
with multi-cell tests). The flag is not subtle; it is a
factor-of-thirty disagreement between the containment design
requirement and the worst-plausible-failure energy. The
calculation is a Fermi estimate at the pre-feasibility level,
not a detailed thermodynamic analysis.

\paragraph{Step 6: the cost-of-rework when the estimate is not
performed.}
The in-service consequences of the missing pre-feasibility
estimate are documented in the public record. The FAA grounded
the global 787 fleet on $16$ January $2013$ and lifted the
grounding on $19$ April $2013$, a three-month grounding that
affected approximately $50$ aircraft in service at the time
\cite{web:ch09t3-faa-787-airworthiness-2013}. Industry
estimates of the lost-revenue cost to the operating airlines
during the grounding cluster around \$1.1~billion in total,
with the per-aircraft revenue at approximately \$60{,}000 per
day and the $50$-aircraft fleet over $90$ days. Boeing's
redesign cost, comprising a new steel-walled containment
enclosure, a ceramic-fibre internal insulation pack, and the
re-certification campaign, has been independently estimated at
approximately \$500~million development cost
\cite{acc:ntsb-aar14-01-787-battery}. The total
cost-of-rework is therefore approximately \$1.6~billion in
external losses plus an unquantified loss of consumer
confidence in the aircraft type.

\paragraph{What the case teaches.}
The pre-feasibility estimate is a single-page calculation. The
inputs are the electrochemistry of the cell, the geometry of
the pack, and the absorptive capacity of the containment box.
The estimate's bottom line is a factor-of-thirty disagreement
between the containment's design assumption and the
worst-plausible-failure energy. The estimate is the
discipline's stopping rule applied to the certification claim:
the claim that single-cell propagation is the design basis is
not supportable in the face of a Fermi estimate that places
multi-cell propagation as a plausible failure mode whose
energy exceeds the containment capacity by a factor of thirty.
The estimate's economic case is the cost-of-rework: an
estimate performed in 2007 at the cost of hours would have
flagged the design gap that cost \$1.6~billion in 2013.

The lesson generalises. The estimation discipline's payoff is
asymmetric in time. An estimate at pre-feasibility costs
hours and saves projects from infeasible architectures. An
estimate at conceptual design costs days and reallocates
budgets. An estimate at detailed design costs weeks and
catches calculation errors. An estimate at certification or
in-service costs months or years and pays the public
cost-of-rework. The reader who has trained the four moves in
this chapter has the technical apparatus; the design habit is
the procedural decision to apply that apparatus at the
earliest phase where it can decide a question.

\subsection{The estimate-then-justify interview: a working
skills assessment}

The working ceiling for the discipline is set by hiring
processes at engineering-led organisations. Senior engineering
interviews at Boeing, Tesla, SpaceX, Google, McKinsey, and the
energy-consulting firms include an estimate-then-justify
exercise as a standard rubric item; the candidate is asked an
order-of-magnitude question, given five minutes of paper time,
and judged on the decomposition, the recorded assumptions, and
the comparison against a published reference value when one
exists \cite{text:ch09t3b-mcdowell-2015,
text:ch09t3b-marn-2014}. The exercise tests what every project
of the design habit tests: whether the candidate carries the
working anchors, whether the candidate decomposes before
calculating, whether the candidate records uncertainty, and
whether the candidate compares against a reference.

Two sample questions, both with published reference values that
permit calibration, work through the exercise.

\begin{estimation}
\textbf{What is the carbon footprint of a one-hour HD video
conference call between two endpoints?}

\smallskip
\textit{Estimate, before reading on. A working interview-grade
question; the published reference is from the International
Energy Agency's $2020$ data-centre and networking briefs.}

\smallskip
The estimate decomposes into the energy consumed by the three
working components (endpoint devices, network transport,
datacentre processing) and the carbon intensity of the grid mix
that supplies the energy.

\textbf{Endpoint power.} Each laptop on the call consumes
$\sim 50 \,\si{\watt}$ during active video. For two endpoints
on a one-hour call, the endpoint energy is $2 \times 50 \times
1 = 100 \,\si{\watt\hour}$.

\textbf{Network transport.} An HD video stream at $2 \,
\si{\mega\bit\per\second}$ over a one-hour call moves
$2 \times 3{,}600 / 8 = 900 \,\si{\mega\byte}$ of data per
direction. The current-as-of-$2020$ working energy intensity of
fixed-line network transport is $\approx 0.06 \,
\si{\kilo\watt\hour\per\giga\byte}$ end-to-end
\cite{data:ch09t3b-iea-data-2020}, so $2 \times 0.9 \times 0.06
= 0.11 \,\si{\kilo\watt\hour} = 110 \,\si{\watt\hour}$ across
both directions. The mobile-network intensity is approximately
$2$-$3 \times$ higher.

\textbf{Datacentre processing.} The hyperscale datacentre
running the conferencing service draws power at $\approx 0.3 \,
\si{\watt}$ per concurrent video stream for the encoding,
mixing, and routing load (a typical $1 \,\si{\mega\watt}$
hyperscale facility hosts $\sim 3 \,\text{M}$ concurrent streams
at peak \cite{data:ch09t3b-iea-data-2020}). For one hour of two
streams the datacentre energy is $2 \times 0.3 \times 1 = 0.6
\,\si{\watt\hour}$, which is negligible at this resolution.

\textbf{Total energy.} The call consumes $\sim 100 + 110 + 1 =
\sim 210 \,\si{\watt\hour}$.

\textbf{Carbon intensity.} The global mean grid carbon
intensity is $\approx 460 \,\text{gCO}_{2}/\text{kWh}$
(current as of $2023$) \cite{data:ch09t3b-ember-2024}; the OECD
mean is $\approx 350 \,\text{gCO}_{2}/\text{kWh}$;
the French mean is $\approx 60 \,\text{gCO}_{2}/\text{kWh}$.
At the global mean, the call's footprint is $0.21 \times 460 =
\sim 100 \,\text{gCO}_{2}$; at the OECD mean,
$\sim 75 \,\text{gCO}_{2}$.

\textbf{Comparison to the published reference.} The IEA's
working $2020$ figure for a one-hour HD video call is between
$\sim 50$ and $\sim 1000 \,\text{gCO}_{2}$ depending on
endpoint type, network, and grid mix, with a midpoint estimate
near $150 \,\text{gCO}_{2}$ for an average global call
\cite{data:ch09t3b-iea-data-2020}. The Fermi estimate of
$\sim 100 \,\text{gCO}_{2}$ at the global mean lands within
the published range and within a factor of two of the
midpoint. The factor that dominates is the network transport
energy; the endpoint power is comparable; the datacentre
contribution is small.

\textbf{Postmortem.} Three sensitivities matter. Wireless
transport raises the network factor by $\sim 3 \times$; LED-backlit
laptop screens versus older models change the endpoint factor by
a factor of two; and grid mix moves the conversion by a factor
of nearly ten between the lowest-carbon and highest-carbon
grids. The candidate who records each factor's range in the
estimate, rather than picking a point value, has captured what
the IEA's published range was capturing.
\end{estimation}

The interview rubric for this question awards full marks for a
recorded decomposition, all four major factors named (endpoint,
network, datacentre, grid intensity), at least one factor's
sensitivity noted, and a comparison against a reference. The
specific point estimate matters less than the discipline; a
candidate whose final number is $50 \,\text{gCO}_{2}$ but
whose decomposition is sound is judged better than a candidate
whose number is $100 \,\text{gCO}_{2}$ but whose
decomposition does not record where the number came from.

\begin{estimation}
\textbf{What is the embodied carbon in a single iPhone 15 Pro
across its full lifecycle?}

\smallskip
\textit{Estimate, before reading on. The published reference is
from Apple's environmental report; the figure is reported per
device.}

\smallskip
The estimate decomposes into manufacturing energy, materials
processing, transport, end-of-life, and use-phase electricity,
with the manufacturing and materials factors dominant for a
smartphone.

\textbf{Manufacturing energy.} An iPhone-class device's
fabrication and assembly draw $\sim 70$ to $90 \,
\si{\kilo\watt\hour}$ of electrical energy (semiconductor
processing for the system-on-chip is the dominant component,
with $\sim 30 \,\si{\kilo\watt\hour}$ for the $A17$-class chip
at a leading-edge node, plus assembly, display, and battery
manufacturing). Taiwan's grid carbon intensity is $\approx 500
\,\text{gCO}_{2}/\text{kWh}$ \cite{data:ch09t3b-ember-2024};
manufacturing emissions are then $\sim 80 \times 0.5 = 40 \,
\text{kg CO}_{2}$.

\textbf{Materials processing.} The aluminium frame contributes
$\sim 7 \,\text{kg CO}_{2}$ for $\sim 0.04 \,\si{\kilo\gram}$
of recycled aluminium at $\sim 200 \,
\text{kg CO}_{2}/\text{t}$ \cite{data:ch09t3b-iai-aluminium-2023}.
The cobalt, lithium, copper, and rare-earth content add
$\sim 5$ to $10 \,\text{kg CO}_{2}$ each from mining and
refining at industry-mean intensities. Glass and ceramics
contribute $\sim 2 \,\text{kg CO}_{2}$. Total materials
processing is $\sim 20 \,\text{kg CO}_{2}$.

\textbf{Transport and end-of-life.} Sea freight from East Asia
to North America at $\sim 0.015 \,\text{kg CO}_{2}/(\text{t}\cdot\text{km})$
over $\sim 10{,}000 \,\si{\kilo\meter}$ for a $0.2 \,\si{\kilo\gram}$
device with packaging at $\sim 0.5 \,\si{\kilo\gram}$ shipped is
$\sim 10^{-2} \,\text{kg CO}_{2}$, negligible at this
resolution. Air freight, however, multiplies by $\sim 100$ to
$\sim 1 \,\text{kg CO}_{2}$. End-of-life recycling
contributes a small negative term, $\sim -1 \,
\text{kg CO}_{2}$.

\textbf{Use-phase.} A four-year service life at $\sim 6 \,
\si{\kilo\watt\hour}/\text{yr}$ of charging at the US grid mean
of $\sim 400 \,\text{gCO}_{2}/\text{kWh}$ contributes
$\sim 10 \,\text{kg CO}_{2}$ over the device life.

\textbf{Total.} Manufacturing $40$ + materials $20$ + transport
$1$ + use $10$ + end-of-life $-1 \approx 70 \,
\text{kg CO}_{2}$.

\textbf{Comparison to the published reference.} Apple's
product environmental report for the iPhone $15$ Pro
($128 \,\si{\giga\byte}$) reports a cradle-to-grave lifecycle
carbon footprint of $66 \,\text{kg CO}_{2}$
\cite{web:ch09t3b-apple-iphone15-2023}, of which $\approx 81\%$
is manufacturing and $\approx 13\%$ is use. The Fermi estimate
of $\sim 70 \,\text{kg CO}_{2}$ lands within $\sim 6\%$
of the published reference. The factor that dominated was
manufacturing energy, in agreement with Apple's reported
breakdown.

\textbf{Postmortem.} The strongest sensitivity is the
manufacturing-energy carbon intensity, which Apple has been
reducing year-over-year through renewable-electricity contracts
at its supply chain; the published iPhone $11$ ($2019$) figure
was $\approx 80 \,\text{kg CO}_{2}$ and the iPhone $14$
Pro ($2022$) was $\approx 65 \,\text{kg CO}_{2}$. The
working trend matters more than any single year's number for an
engineer deciding between repair and replacement.
\end{estimation}

The two estimation blocks illustrate the working interview
rubric and the working engineering rubric simultaneously: a
candidate who reaches a number within a factor of two of the
reference by a recorded decomposition with explicit sensitivities
has done what the engineering canon now requires the candidate
to do. The reader who has practised the discipline on the
chapter's earlier estimates is now equipped to walk into any
estimate-then-justify interview at this level.

\subsection{The estimate as a project-portfolio filter}

A design organisation that funds many projects faces a
portfolio-level estimation question that no single project
asks. Across the organisation's queue of candidate projects,
which projects are worth allocating engineering effort to, and
which are not? The portfolio-level estimate is the same
Fermi-style decomposition the chapter has trained, applied to
each candidate's feasibility, cost, and expected value, then
ranked.

The discipline at portfolio level is the same discipline at the
project level: the estimate is recorded before the detailed
analysis, the estimates are compared across projects on a
consistent basis, and the postmortem catches estimates that
biased the portfolio's funding decisions. The portfolio-level
estimate is also where the planning fallacy of
section~9.5 leaves its largest residue; the estimates produced
by the engineers who will execute the project are biased low
by a factor of $1.5$ to $2$, and the portfolio's expected-cost
ranking inherits that bias. Reference-class forecasting at the
portfolio level, anchored to the organisation's empirical
distribution of past project outcomes, is the discipline that
corrects for the bias.

The Boeing 787 case described above is a project-level miss;
the same mechanism shows up at portfolio level when an
organisation's estimates of programme cost, schedule, and
technical risk are produced by the teams that will execute the
programmes. The portfolio-level discipline is the same: record
the estimates, compare to the reference class, and require
reconciliation when the team estimate disagrees with the
reference class by more than the team's stated uncertainty.
The discipline does not eliminate the planning fallacy; the
discipline names it and produces a record against which the
realised cost and schedule can later be compared.

\section{Estimation under deep uncertainty}\pathtag{core}

The estimates of section 9.1 through section 9.4 assume that each
factor has a defensible nominal value within a factor of two or
three. A separate class of estimation problem arises when one or
more factors are themselves only known within an order of
magnitude, or when the underlying physics or population is itself
uncertain. We call this \engterm{deep uncertainty}, after the
convention in risk-analysis literature
\cite{paper:lempert-2003-shaping-future,paper:walker-2003-deep-uncertainty},
and treat it with a different set of moves.

The working risk-analysis literature distinguishes four levels of
uncertainty. \engterm{Level 1} ("a clear-enough future"): a
single trajectory of plausible outcomes; standard probability
analysis applies. \engterm{Level 2} ("alternate futures"): a few
discrete scenarios with known probabilities; expected-value
optimisation applies. \engterm{Level 3} ("a multiplicity of
plausible futures"): many scenarios, probabilities unknown or
contested; scenario analysis and robust decision-making apply.
\engterm{Level 4} (deep uncertainty proper, "the unknown
future"): the set of scenarios is itself unknown; ignorance about
the structure of the problem rules out scenario analysis. The
working engineer encounters Level 3 and Level 4 in climate
adaptation, supply-chain disruption planning, long-horizon
infrastructure investment, and novel-technology risk assessment.

The deep-uncertainty literature has developed working methods
that share a common structure: identify the decision, identify
the candidate actions, identify the futures or scenarios, and
choose the action whose worst-case performance across the
scenario set is acceptable rather than the action whose
expected-value performance is optimal. The methods differ in
how they generate the scenario set and how they aggregate
performance across scenarios.

\engterm{Scenario analysis} (Royal Dutch Shell since the 1970s):
identify $3$-$5$ distinct internally-consistent futures (the
canonical examples are Shell's "Mont Fleur" South Africa
transition scenarios, $1991$-$1992$, and the IPCC's SSP1 through
SSP5 socioeconomic pathways for climate modelling) and stress-
test each candidate action against all scenarios. The decision
chosen performs adequately across the scenario set even if no
scenario carries assigned probability. Working tools: Shell's
in-house Strategic Group methodology, the Global Business
Network's "$2 \times 2$ matrix" scenario-axis construction,
and the World Bank's CCDR (Country Climate and Development
Report) scenarios for climate-stressed national investment
planning.

\engterm{Robust decision-making (RDM)} (Lempert, Popper, and
Bankes at RAND from the late 1990s
\cite{paper:lempert-2003-shaping-future}): use computer-simulation
ensembles of $10^{3}$ to $10^{5}$ futures sampled from
deep-uncertainty parameter distributions, then identify the
\engterm{scenarios that distinguish among candidate strategies}
through statistical clustering. The chosen strategy is robust
(performs adequately across most futures) and adaptive (signals
when the actual future is diverging from the assumed scenario
set). Working software: RAND's Robust Decision-Making toolkit;
Carnegie Mellon's EOLAB framework; Lawrence Livermore National
Laboratory's RDMtoolbox in MATLAB. Applications: California
water management (the $2009$ California Water Plan), Louisiana
coastal master plan, US Department of Defense long-range
strategic planning.

\engterm{Info-gap decision theory} (Yakov Ben-Haim
\cite{text:ch09-ben-haim-2006-info-gap}): defines a horizon of
uncertainty $\alpha$ around a best-estimate parameter value and
asks "for what range of $\alpha$ does each candidate action
deliver acceptable performance?". The action with the largest
robustness $\alpha$ is preferred. Info-gap distinguishes itself
from RDM by not requiring a probability distribution or even a
scenario set on the uncertain parameters; the only input is the
best estimate and a definition of "acceptable performance".
Applications: river-flood embankment design under climate
uncertainty, biopharmaceutical inventory planning under
demand uncertainty, defence-system architecture under threat
uncertainty.

\engterm{Decision-making under deep uncertainty (DMDU)
community} (formalised 2015 onward through the Society for
Decision Making under Deep Uncertainty, SDMDU): a synthesis of
the above methods plus dynamic adaptive policy pathways (DAPP),
exploratory modelling and analysis (EMA), and multi-objective
robust decision-making (MORDM). The community publishes through
the journal \textit{Environmental Modelling \& Software} and the
annual SDMDU conference; the working library is the Exploratory
Modeling Workbench (EMA Workbench, written in Python by Jan
Kwakkel at TU Delft, BSD licence, available on PyPI as
\texttt{ema\_workbench}). Applications: Netherlands Delta
Programme flood defence, Singapore PUB's national water plan,
New York City Office of Climate \& Environmental Justice
infrastructure-investment master plan.

\subsection{When the factor's uncertainty is the answer}

For \emph{the total mass of microplastics in the world's oceans},
the answer depends on particle-size cutoff, sampling depth, polymer
identification method, and whether the estimate concerns surface
waters, the upper water column, sediments, or the whole ocean.
Published estimates therefore span orders of magnitude rather than
a neat factor-of-three band. A defensible estimate must state the
compartment, size range, sampling assumption, and source set before
it reports a range \cite{paper:zhao2025-ocean-microplastics}.

A deep-uncertainty estimate reports the range as the answer. The
discipline is to identify the upper and lower bounds defensible
from the available evidence, state the range explicitly, and
identify the factor whose uncertainty contributes most to the
range's width. A point estimate without its range misrepresents the
problem.

\subsection{Bounding from above and below}

Two complementary moves are available when no nominal value is
defensible.

\engterm{Upper bound}. The reader constructs the largest plausible
value for each factor, multiplies, and obtains an upper bound on the
result. Where a physical conservation law constrains the answer
(energy, mass, momentum cannot exceed available reserves), the
conservation law sets a tighter upper bound than the multiplicative
estimate. For \emph{total metabolic energy a single human expends in their
lifetime}, an upper bound is set by metabolic power times lifespan.
A rough adult metabolic power of $100 \,\si{\watt}$ sustained for
$80$ years gives $100 \times 80 \times 365.25 \times 86400 \approx
2.5 \times 10^{11} \,\si{\joule}$; a long-lived, active upper bound
is therefore closer to $3 \times 10^{11}$ to $4 \times 10^{11}
\,\si{\joule}$ \cite{text:national-academies-energy-1989}. The
conservation argument is hard; the previous power-of-ten must still
be checked.

\engterm{Lower bound}. The reader constructs the smallest plausible
value for each factor, multiplies, and obtains a lower bound on the
result. Lower bounds are often easier than upper bounds because
existence arguments suffice: the existence of a single working
example proves that the quantity is at least as large as that
example.

The two bounds together identify the range within which the answer
must lie. When the bounds are within a factor of three of each
other, the problem reduces to ordinary Fermi estimation. When the
bounds span more than an order of magnitude, the problem is a
deep-uncertainty problem and the bounds themselves are the answer.
\Cref{fig:vol01:ch09:bounds-narrowing} shows the move as a process:
each factor pinned to a defensible value pulls the upper bound down and
the lower bound up, and the reader watches whether the two sides close
below the factor-of-three target or refuse to. The factor whose two
sides remain farthest apart after every other factor is pinned is the
one the postmortem names as the width's source, and it is the factor
whose measurement, where a measurement is possible, would convert the
deep-uncertainty problem into an ordinary one.

\input{volumes/01-quantity/ch09-discipline-of-estimation/figures/fig-bounds-narrowing.tex}

\paragraph{Worked instance: expert-judgment aggregation under
deep uncertainty.}
A working laboratory needs an estimate of an unmeasurable quantity:
the probability that a buried-waste container at a particular site
will leak within $1000$ years. No directly relevant accelerated-test
data exist; the corrosion timescales exceed any feasible
experiment, and the geological signals at the site are themselves
uncertain. The discipline that the probabilistic-risk-analysis
literature has built for these cases is structured expert
elicitation \cite{ch09-paper:bedford-cooke-2001,
paper:ch09-cooke-expert-2008}. The reader runs the following
procedure. (i) Identify a panel of $5$ to $10$ experts whose
working knowledge spans the relevant disciplines (corrosion
chemistry, hydrogeology, container engineering). (ii) For each
expert, elicit $5\,\%$, $50\,\%$, and $95\,\%$ quantiles on the
target quantity individually, with no inter-expert discussion. (iii)
Calibrate each expert against a set of seed questions whose answers
are known to the elicitor but not to the experts; the seed
questions test how well each expert's stated uncertainty matches
their realised accuracy. (iv) Aggregate the experts' distributions
weighted by their calibration scores on the seed questions, not by
seniority. The aggregated distribution is the deep-uncertainty
estimate. The Cooke method's published case studies report that
calibration-weighted aggregation produces tighter, better-calibrated
distributions than equal-weighted aggregation, because the weights
deweight overconfident experts whose stated $90\,\%$ intervals
miss the seed answers more often than $10\,\%$ of the time
\cite{paper:ch09-cooke-expert-2008}. The procedure is the working
laboratory's substitute for the experiment that cannot be run;
the discipline is to record the elicitation and the
calibration-weighted aggregation as the estimate, rather than to
present a single number as if it had emerged from a calculation.

\subsection{Megaprojects and the planning fallacy}

Project cost and schedule are the largest single class of practical
estimation failures in engineering, larger by far in dollar terms
than the technical accidents named elsewhere in the chapter.
Flyvbjerg's study of $258$ transport infrastructure projects over a
$70$-year period found cost overruns in $86\,\%$ of projects, with a
mean overrun of $28\,\%$ and a one-sigma spread that reached
$65\,\%$; the study also documented that the overrun statistics had
not improved across the decades the data covered
\cite{paper:ch09-flyvbjerg-cost-overrun-2002,text:ch09-flyvbjerg-megaprojects-2003}.
Kahneman and Tversky named the underlying mechanism the
\emph{planning fallacy}: a systematic tendency of forecasters to
underestimate the time and cost of their own projects, even when
they correctly estimate the time and cost of comparable past
projects done by other people
\cite{paper:ch09-kahneman-thinking-1979}.

Two illustrative cases. The Sydney Opera House was approved in
$1957$ on a cost estimate of AUD~$7$ million and a four-year
construction schedule. The building opened in $1973$, ten years
late, at a final cost of AUD~$102$ million, a factor-of-fourteen
overrun in nominal currency \cite{paper:ch09-bia-1972-sydney}. The
Boston Central Artery (the Big Dig) was approved in $1985$ on a
cost estimate of \$2.8 billion. The tunnel system completed in
$2007$ at a cost of approximately \$14.8 billion, a factor-of-five
overrun in nominal currency and a factor-of-three overrun in
constant dollars \cite{text:ch09-flyvbjerg-megaprojects-2003}.

The estimation lesson is that cost estimates produced inside the
organisation responsible for the project are systematically
biased; the defence is what Flyvbjerg calls \engterm{reference
class forecasting}: anchor the new project's cost on the empirical
cost distribution of similar past projects, then adjust for what
makes this project different. The discipline is the same
estimate-before-calculate move the chapter has been training, with
the reference class supplying the anchor. A planning team that
records the reference-class estimate before producing the
project-specific estimate has a defence against the planning
fallacy that no amount of internal analysis supplies.

The mechanism applies in the small as much as in the large. A
software-feature estimate produced by the engineer who will build
the feature is, on the empirical record of agile-development
teams, biased low by a factor of $1.5$ to $2$ for the median task,
with a long upper tail dominated by tasks whose scope was
underestimated during planning
\cite{ch09-paper:bent-2018-overruns}. The reference-class defence
works at this scale too. The planning team records the average
ratio of actual to estimated effort across the team's last $50$
to $100$ completed tasks; the new task's expected effort is the
engineer's bare estimate multiplied by that ratio. The procedure
is simple, the discipline is to apply it before the schedule is
committed to. A team that applies the correction once per
quarter, and that records its current empirical ratio in the
planning system, has converted the planning fallacy from a
recurring loss into a measurable, declining one.

A second pattern in the megaproject record is what Albert
Hirschman called the \emph{Hiding Hand}: organisations
underestimate both the difficulty of a project and the
creativity with which they will overcome the difficulty, and the
two errors partially cancel
\cite{ch09-paper:bent-2018-overruns}. The pattern is consoling
but defensible only in retrospect, and the cancellation rate is
neither $100\,\%$ nor reliable across projects. A planning
discipline that assumes the Hiding Hand will rescue the
underestimated effort is, on the empirical megaproject record, a
planning discipline that produces failed projects at a rate the
historical record can quantify. The reference-class defence is
the chapter's recommended substitute: use the empirical
distribution of past outcomes, not the consoling story about how
this project will be different.

\subsection{The MacKay convention: estimate, then revisit}

David MacKay's \emph{Sustainable Energy: Without the Hot Air}
\cite{gen:mackay2009} is a sustained worked example of deep-
uncertainty estimation in the energy domain. MacKay's discipline,
adopted here, is to give the best available point estimate, with
the underlying assumptions explicit, and to invite the reader to
revisit the estimate as new data become available. The estimate is
dated, so a later reader knows when to revisit it. The discipline of
dating estimates is what lets a reader, ten years later, see whether
the estimate has aged.

\subsection{A second canonical estimate: world plastic production}

The second canonical estimate of the chapter exercises the same four
moves on a different domain. The question is not deep-uncertainty in
the technical sense, since a credible industry total exists, but the
two principal factors (per-capita plastic use and the structural
correction for income disparity) carry wider working uncertainties
than the piano-tuner factors. The estimate is also one whose ground
truth has migrated in the last two decades, which makes the
date-the-estimate discipline visible.

\begin{estimation}
\textbf{What is the total mass of plastic produced worldwide in one
year, in tonnes, current as of 2022?}

\smallskip
\textit{Estimate, before reading on. The factors that come up are
world population, per-capita annual plastic use, and a structural
correction for income disparity in consumption.}

\smallskip
The world population in 2022 was $8.0 \times 10^{9}$, current per
the U.N.\ World Population Prospects. The mean annual plastic
consumption per capita is the harder factor. In a high-income
country a working anchor is $100 \,\si{\kilogram}$ per person per
year (packaging, fibres, durable goods, construction). In a
low-income country the per-capita figure is closer to
$10 \,\si{\kilogram}$ per person per year. The income-weighted
world mean sits between, around $50 \,\si{\kilogram}$ per person
per year. The product is then
\[
8.0 \times 10^{9} \,\text{persons}
\times 50 \,\si{\kilo\gram\per\text{person\,yr}}
\approx 4 \times 10^{11} \,\si{\kilo\gram}
= 4 \times 10^{8} \,\si{\tonne}.
\]
The defensible range, given a per-capita factor uncertainty of about
$1.5\times$ and a structural-correction uncertainty of about
$1.2\times$, is approximately $2.5 \times 10^{8}$ to $5.5 \times
10^{8}$ tonnes.

Plastics Europe's \emph{Plastics, the Facts 2023} reports global
plastics production of approximately $400 \,\si{\mega\tonne}$ for
2022 \cite{data:ch09-plastics-europe-2023}. The estimate lands on
the published figure to within $10\,\%$.

The postmortem is short: the per-capita-use factor was the
dominant uncertainty; the published number is dominated by
high-income-country and Chinese consumption, and the income-weighted
mean of $50 \,\si{\kilo\gram}$ per person per year was the lever
the estimate turned on. A reader who later learns that
high-income-country per-capita plastic use is closer to
$140 \,\si{\kilo\gram}$ per person per year (and that production
includes non-final-consumption use such as industrial plastics) can
narrow the estimate by about $20\,\%$. The estimate also dates
quickly: the 2022 figure may already be replaced by a 2025 figure
that includes faster growth in middle-income countries. The
estimate's defensibility lives in being able to recompute when the
inputs change.
\end{estimation}

\subsection{Worst-case versus expected-case}

The deep-uncertainty estimate is sometimes asked for its worst case
rather than its expected case. The two are different objects. The
worst-case estimate is the upper bound described above; it is
appropriate when the consequence of underestimation is severe. The
expected-case estimate is the central value of the distribution; it
is appropriate for budgeting, planning, and ordinary engineering
decisions.

The reader should not collapse the two. A worst-case fuel
consumption used to size a fuel tank, reported as the expected
fuel consumption, leads to a tank designed for a regime that
rarely occurs and to a vehicle that carries unnecessary weight on
every flight. An expected-case structural load used to size a
beam, reported as the worst-case load, leads to a beam that fails
when the actual worst case arrives. The discipline is to name the
case the estimate represents.

A third case-type is the \engterm{robust-design} estimate, which
chooses an action whose performance is acceptable across the
plausible range of cases without committing to any single one of
them as the true case. The robust-design estimate is the right
form when the cost of being wrong about which case is true is
larger than the cost of a slight loss of efficiency under the
true case. The trade is explicit: the engineer accepts a degree
of suboptimality under each possible case in exchange for a
bounded performance across all cases. Coastal-protection design,
which must protect against both extreme storms and slow sea-level
rise without committing in advance to which scenario dominates,
is the recurring civil-engineering instance
\cite{ch09-paper:hallegatte-2009}.

\subsection{Worked case: a climate-stressed design load}

The deep-uncertainty estimate this chapter has been describing in
prose has a long worked instance in current civil engineering: the
50-year-return-period rooftop snow load on a steel-frame warehouse
under projected climate change. The case is chosen for its
combination of three deep-uncertainty features. The historical
record sets a baseline. The forward-looking projection sits in a
distribution whose width is itself uncertain. And the structural
decision (the rooftop snow load to size the trusses against) must
be a single number, even though the underlying quantity is a
random variable whose distribution is not stationary.

\paragraph{Step 1: state the design quantity and the inherited
convention.}
ASCE 7-22 \cite{ch09-std:asce-7-22} specifies design ground snow
loads for the contiguous United States by mapped 50-year mean
recurrence interval (MRI) values, with the design rooftop snow
load derived from the ground snow load through an exposure
coefficient $C_{e}$, a thermal coefficient $C_{t}$, and an
importance factor $I_{s}$. For a heated warehouse in Buffalo, NY,
the ground snow load is $p_{g} = 2.4 \,\si{\kilo\pascal}$ ($50 \,
\text{lbf}/\text{ft}^{2}$) per the mapped value; with $C_{e} =
1.0$, $C_{t} = 1.0$, $I_{s} = 1.0$, the flat-roof snow load is
$p_{f} = 0.7 \, p_{g} = 1.7 \,\si{\kilo\pascal}$.

\paragraph{Step 2: estimate the parameter uncertainty in the
ground snow load itself.}
The mapped value of $p_{g}$ is the 50-year mean recurrence
interval estimate from the historical station-record analysis
that ASCE 7-22 carries forward. The station-record sample size is
of order $50$ to $80$ winters depending on the station. The
sample-derived standard error of the $50$-year quantile under a
generalised extreme value (GEV) model is approximately $\sigma /
\sqrt{n}$ times a tail-quantile amplification factor of order
$3$ to $5$. For the Buffalo station the published one-sigma
uncertainty on $p_{g}$ is approximately $\pm 0.4 \,\si{\kilo
\pascal}$, or $\pm 17\,\%$ of the central value, before any
climate adjustment.

\paragraph{Step 3: estimate the climate-change shift.}
The forward-looking question is whether the 50-year MRI value
itself shifts as the climate warms. The IPCC AR6 Working Group I
reports projected changes in extreme precipitation and snow water
equivalent at the regional level
\cite{ch09-std:ipcc-ar6-wg1}. The relevant signal for the U.S.
Northeast is a projected decrease in mean snow depth (warmer
winters) combined with an increased variance in extreme events
(stronger storms in the remaining cold periods). For a warming
scenario at the median of the AR6 SSP2-4.5 emissions pathway, the
projected mid-century shift in the 50-year ground snow load for
the U.S. Northeast is on the order of $-10\,\%$ to $+15\,\%$
relative to the late-twentieth-century baseline, with the sign
unresolved at the $50$-year-MRI tail. The bounds are deeper than
ordinary statistical bounds because the underlying climate model
ensemble disagrees about the direction of the shift at this MRI.

\paragraph{Step 4: combine the bounds into a deep-uncertainty
range.}
The combined range on $p_{g}$, mid-century, is the statistical
uncertainty composed with the climate-shift range. Statistical
one-sigma: $\pm 17\,\%$. Climate-shift bounds: $-10\,\%$ to
$+15\,\%$, treated as a uniform interval whose standard deviation
is $25\,\% / \sqrt{12} \approx 7\,\%$. Combined in quadrature
under the independence assumption (a defensible first cut):
$\sqrt{17^{2} + 7^{2}} \approx 18\,\%$. The implied combined
one-sigma is $\pm 0.43 \,\si{\kilo\pascal}$ on a central value
that is itself shifted by the climate signal. The central value
could plausibly be anywhere from $0.9 \times 2.4 = 2.16 \,\si
{\kilo\pascal}$ to $1.15 \times 2.4 = 2.76 \,\si{\kilo\pascal}$,
with a $\pm 0.4 \,\si{\kilo\pascal}$ band around each.

\paragraph{Step 5: convert the range into a robust design value.}
The robust-decision-making literature recommends, for cases like
this, that the design value be chosen by minimising regret across
the plausible range rather than by selecting the central value
\cite{ch09-paper:hallegatte-2009}. The candidate decision rules
are: (a) design to the central value, $1.7 \,\si{\kilo\pascal}$
roof load, accepting an elevated failure probability if the
upper-bound climate scenario realises; (b) design to the
upper-bound central value plus one sigma, $1.15 \times 2.4 + 0.4
= 3.16 \,\si{\kilo\pascal}$ ground load, $0.7 \times 3.16 = 2.2 \,
\si{\kilo\pascal}$ roof load, paying a $30\,\%$ structural
premium against a tail risk that may not materialise; (c) design
to the central value and budget for retrofit if the upper-bound
scenario materialises.

The retrofit-budgeted option (c) is often the lowest-cost
decision in expectation, but it depends on the building's
intended life and on whether the upper-bound climate scenario can
be detected before structural failure occurs. For a warehouse
with a $50$-year design life, the central-design-plus-monitoring
option dominates by expected cost; for a hospital with a $100$-
year design life and a low tolerance for structural failure
during a snow event, the upper-bound design option (b) is
defensible despite the premium.

\paragraph{Step 6: report the design value with its range and the
decision rule that produced it.}
The defensible report names the design value, the bounds on the
underlying quantity, and the decision rule. ``Designed to $1.7 \,
\si{\kilo\pascal}$ roof snow load, central value of the
ASCE 7-22 mapped ground load at the site. The combined 50-year
MRI ground load, including climate-projection uncertainty, lies
in $2.16 \pm 0.43$ to $2.76 \pm 0.43 \,\si{\kilo\pascal}$ at
the one-sigma level, current as of 2024 climate projections. The
design rule is central-value design with provision for
monitoring; the structural budget for a retrofit to a $20\,\%$
heavier load is $X\,\%$ of as-built cost.'' The discipline is to
report all three: the value, the range, the rule. A reader given
only the value cannot tell whether the value reflects a central
estimate or an upper bound, cannot evaluate the residual risk,
and cannot decide whether monitoring is adequate.

The general pattern recurs whenever an engineering decision must
fix a single number against a quantity whose distribution is
itself uncertain. Coastal-protection design heights, hydrologic
design floods, wind-load design pressures, and lifetime-fatigue
load spectra are the recurring instances. The discipline names
the deep-uncertainty range, identifies the decision rule, and
records both alongside the chosen number, so a later engineer
can revisit the choice when the underlying climate or
hydrological projections are revised.

\paragraph{What the case adds to the chapter's apparatus.}
The four-moves estimation environment is a single block on a
single page; the climate-stressed design-load case is the same
discipline run across three sources of uncertainty and a
decision rule. The general lesson is that the moves scale. The
question, in deep-uncertainty problems, includes a specification
of the case the estimate represents (worst, expected, robust). The
estimate, in deep-uncertainty problems, is a bounded range plus a
decision rule rather than a point. The comparison, in
deep-uncertainty problems, is against the cost of being wrong in
each direction, not against a single ground-truth value. The
postmortem, in deep-uncertainty problems, identifies which of the
contributing uncertainties dominated the range's width and
schedules the work that would tighten it. The reader who has
internalised the four moves for ordinary Fermi problems has the
apparatus for the deep-uncertainty version with a small extension
to each move.

The historical sketch on the discipline closes here. The
deep-uncertainty form is the modern descendant of the worst-case
engineering tradition that produced steam-pressure safety
factors of $3$ to $10$ in the nineteenth century, the load-and-
resistance-factor codes of the twentieth, and the structured
expert-elicitation methods of the late twentieth and twenty-first.
Each generation has built apparatus to handle a wider class of
unknowns. The reader at the close of Volume I is reading the
current state of the art on Fermi-style estimation. The estimate's
honest form is a number with its range, the decision rule that
produced it, and the date on which it was current; a reader who
records all three has the working apparatus that the chapter has
been training.

\subsection{Compound effects: estimating low-probability,
high-consequence events}

Deep-uncertainty problems often involve compound effects: events
that are individually low-probability but jointly catastrophic when
they coincide. The estimation discipline at this end of the
spectrum has to address two questions the ordinary Fermi machinery
does not answer. How does an engineer assign a probability to an
event that has never occurred in the historical record? And how
does an engineer combine the consequence of such an event with the
probability to produce a working risk estimate?

The Fukushima Daiichi case is the chapter's worked example of the
compound-effect class. The pre-2011 design-basis tsunami for the
Daiichi site, set in the early 1970s when the plant was built,
was approximately $5.7 \,\si{\meter}$ above mean sea level, based
on the maximum recorded historical tsunami at the site
\cite{paper:ch09t3-toda-tsuruoka-2012}. The deterministic estimate
took the largest recorded event in the regional historical record
(the $1896$ Sanriku tsunami) as the worst plausible case. The
probabilistic estimates carried out by TEPCO between $2000$ and
$2009$, however, identified a non-trivial probability of waves
substantially larger than the design basis when the geological
record (including the $869$ Jogan tsunami inferred from sediment
deposits) was included; those probabilistic estimates indicated a
$\geq 10 \,\si{\meter}$ tsunami at $50$ to $100$-year mean
recurrence interval would be defensible but did not lead to
operational changes \cite{acc:tepco-fukushima-2012,
paper:ch09t3-toda-tsuruoka-2012}.

The $11$ March $2011$ tsunami reached approximately $13 \,\si{\meter}$
at the site, overtopping the seawall by more than a factor of
two and submerging the backup diesel generators located at the
low-elevation seawater intake \cite{acc:tepco-fukushima-2012}.
The compound effect was the simultaneous loss of offsite power
(from the earthquake) and the loss of all backup power (from the
tsunami); the redundancy that had been engineered into the backup
system was defeated by a common-mode failure that the
deterministic design basis had not considered. The realised
cost of the wrong tail is documented: the Japanese government's
$2023$ decommissioning and compensation cost estimate stood at
approximately $\textyen 21.5$~trillion, or about \$190~billion at
contemporary exchange rates \cite{web:ch09t3-tepco-decommissioning-2023},
not counting the indirect costs of the loss of the
$\sim 47 \,\text{GW}$ of Japanese nuclear capacity that followed.

The estimation lesson is structural. A deterministic
worst-historical-case estimate is a defensible lower bound on
the design hazard but is not a defensible upper bound. The
probabilistic estimate, drawing on the geological and
paleoseismic record rather than only on the instrumental record,
identified the tail that the deterministic estimate had ruled
out. The compound-effect estimate, combining the tsunami
probability with the joint loss-of-power consequence, produces
a risk number that a deterministic estimate alone cannot. The
discipline at this end of the deep-uncertainty spectrum is to
perform both kinds of estimate and to record both ranges, then
to choose the design value through the robust-decision-making
move described above. A design value chosen against only the
deterministic estimate inherits the deterministic estimate's
silent tail assumption; a robust-design value chosen against
the probabilistic-and-deterministic envelope is honest about
the tail.

\subsection{Black-swan risk and position sizing}

Taleb's $2007$ framing of the \engterm{black swan} is the
working term for the class of events outside the historical
record whose probability and consequence are both
unestimable from in-sample data
\cite{text:ch09t3-taleb-2007}. The framing applies to
engineering whenever the design domain extends into a regime
that the available data do not cover: novel materials at
extreme temperatures, software at unprecedented scale,
financial leverage at multi-sigma deviations, and so on. The
estimation discipline at this regime is not to estimate the
probability at all; it is to size the engineered system's
exposure so that survival is possible even if the event
occurs.

The Long-Term Capital Management collapse of $1998$ is the
working financial-engineering instance. LTCM had been founded
in $1994$ on a fixed-income arbitrage strategy whose
historical Sharpe ratio over the prior five years had been
exceptional; the strategy's risk model was a multivariate
Gaussian fit to the historical correlation structure of
sovereign-bond spreads. The Russian sovereign default of
August $1998$ produced a coordinated movement in spreads that
the historical sample had no analogue for: a movement
estimated by the fund's risk model at approximately $10$~sigma,
that is, a per-event probability of $\sim 10^{-23}$ under the
Gaussian model
\cite{text:ch09t3-lowenstein-2000,paper:ch09t3-mackenzie-2003}.
The realised consequence was a fund equity loss of \$4.6
billion over six weeks and an emergency recapitalisation
arranged by the Federal Reserve to avoid systemic contagion.

The estimation lesson is the same as Fukushima at the
financial scale. A risk model fit to in-sample data
underestimates the probability of out-of-sample movements;
the probability cannot be honestly estimated from data that
does not contain the events of interest. The discipline at
this regime is to compute the position size against the
worst-plausible-realisation rather than against the
historical distribution, and to record the position-sizing
rule alongside the in-sample risk number. LTCM's position
sizes had been computed against a $4$-sigma worst case in
the Gaussian model; the surviving discipline in the
post-$1998$ financial-engineering literature is to compute
position sizes against a regime-shift worst case that the
historical model cannot calibrate. The procedural defence is
explicit: the risk number is recorded with a note identifying
the regime in which it applies, and the position size is
chosen so that the fund or the engineered system survives a
regime shift, not so that it optimises performance under the
calibrated regime.

The principle generalises. A novel-technology risk assessment
that uses only in-domain historical data underestimates the
tail. A reliability prediction for a new alloy at extreme
temperature underestimates the probability of phase
transitions outside the calibrated range. A control-system
stability margin computed against the linearised model at the
operating point underestimates the probability of mode
transitions to a regime the linearisation does not capture.
The estimation discipline's answer in each case is the same:
record the in-sample estimate, identify the regime in which
it applies, and size the engineered system's exposure so that
out-of-regime events do not lead to system loss.

\subsection{Probabilistic risk assessment as engineering apparatus}

The probabilistic risk assessment (PRA) framework, formalised
in the U.S.\ Nuclear Regulatory Commission's NUREG-1150
reactor-safety study of $1990$, is the engineering apparatus
that handles a structured version of the compound-effect
problem \cite{paper:ch09t3-nureg-1150}. PRA constructs an
event tree that enumerates the sequences from an initiating
event (loss of offsite power, large break loss-of-coolant,
seismic ground motion above design basis) through the
intervening barriers (emergency diesel start, high-pressure
injection, containment isolation) to one of several end
states (core damage, large early release of radioactivity, no
release). Each branch of the tree carries a conditional
probability; the tree's quantification produces the working
risk numbers that the framework is designed to compute.

Two end-state risk metrics dominate Gen II and Gen III
reactor PRA. The \engterm{core damage frequency} (CDF) is the
per-reactor-year probability of core melt, computed by summing
the probabilities of all sequences whose end state is core
damage. The \engterm{large early release frequency} (LERF) is
the per-reactor-year probability of a containment breach with
substantial radioactive release before evacuation of the
surrounding population can be completed. For Gen II reactors
the CDF target set by the NRC is approximately $10^{-4}$ per
reactor-year; the LERF target is approximately $10^{-5}$ per
reactor-year. For Gen III reactor designs (AP1000, EPR,
ABWR), the design targets are an order of magnitude lower:
CDF $\sim 10^{-5}$, LERF $\sim 10^{-6}$ per reactor-year
\cite{paper:ch09t3-nureg-1150}. The working software for PRA
in U.S.\ practice is SAPHIRE, maintained by Idaho National
Laboratory under NRC sponsorship
\cite{paper:ch09t3-saphire}; complementary tools include the
EPRI-developed RISKMAN and the Sandia-developed MELCOR
sequence-modelling code.

PRA's estimation discipline is the chapter's discipline at
scale. Each branch probability is itself a Fermi estimate
(from component-test data, in-service experience, expert
judgement, or a combination); the sequence probabilities are
products of branch probabilities, with the log-quadrature
combined uncertainty that section~9.2.3 introduced. The PRA
result is a probability number with an associated uncertainty
range; in the working NRC reporting standard, the 5th, 50th,
and 95th percentiles of the CDF and LERF are reported
alongside the point estimate, and the dominant contributors
to the uncertainty range are identified explicitly. The
discipline at PRA level is the chapter's discipline at
project level multiplied across hundreds of branches.

The estimation lesson PRA teaches is that the deep-uncertainty
problem can be made structured even when the underlying
probabilities are not directly measurable. The event-tree
structure forces every analyst's assumptions into the open;
the per-branch quantification forces every probability
estimate to have a documented source; the uncertainty
propagation forces every result to carry its range. A PRA
report that meets the NRC's documentation standard is a
worked record of the chapter's four moves applied at scale.
The reader who has practised the four moves on fifty Fermi
problems at the end of this chapter has the working
apparatus that PRA uses; the difference is the count of
branches and the documentation discipline, not the underlying
estimation move.

\subsection{When the tail-risk estimate mattered: New Madrid
seismic design loads}

The estimation discipline can produce a defensible tail-risk
number even when the tail event has not occurred in the
instrumental record. The Central United States seismic
design loads documented in the International Building Code
Section~1613 are a working example
\cite{std:ch09t3-ibc-2018}. The IBC adopts the U.S.\
Geological Survey's probabilistic seismic-hazard maps as the
design-basis ground motion for buildings; the maps reflect a
$2\,\%$ probability of exceedance in $50$ years, equivalent
to a $\approx 2500$-year mean recurrence interval.

For Memphis, Tennessee, located approximately $90 \,\si{\kilo
\meter}$ from the New Madrid Seismic Zone (NMSZ), the
design-basis spectral acceleration at $1$~second period is
approximately $0.5 \,\text{g}$ \cite{std:ch09t3-ibc-2018}.
The number is a working design load that drives Memphis
building costs upward by an estimated $5\,\%$ to $10\,\%$
relative to a no-seismic-load city of the same size. The
instrumental seismic record at Memphis is sparse; the
USGS-recorded maximum-ever ground motion at the site is well
below the $0.5 \,\text{g}$ design number. The tail-risk
estimate that produced the design number is built on the
paleoseismic record \cite{text:ch09t3-tuttle-2002}: the
$1811$-$1812$ NMSZ earthquakes, three magnitude-7 events in
three months, produced ground shaking at Memphis that the
paleoseismic record (sand-blow deposits, deformation
features) identifies as the realised tail event of the
recurrence-interval distribution. The $1811$-$1812$ events
are the calibration data for the $0.5 \,\text{g}$ design
number; they are the ``$200$-year ground truth'' against which
the probabilistic estimate is anchored.

The case teaches the inverse of the LTCM lesson. Where the
in-sample data do not contain the tail event, the engineer
who can extend the sampling window through the geological,
paleoseismic, or historical record acquires a defensible tail
estimate where the instrumental record alone cannot supply
one. The discipline is to identify the extended data source,
document the inference from the proxy data to the tail
probability, and record the inferred tail estimate alongside
the in-sample estimate. Memphis's $0.5 \,\text{g}$ design
ground motion is the worked instance: a defensible tail
number derived from a proxy data source whose calibration
against the instrumental record forces the inference to be
explicit. A reader who returns to the deep-uncertainty
problem after a Volume X case study can apply the same
move: identify the proxy record that the in-sample data
omit, document the inference, record the extended estimate.

\section{The estimation environment: a formal apparatus}\pathtag{core}

The reader has by now used the \verb|estimation| environment in every
previous chapter of this volume. Five early examples are the Boeing 747
maximum take-off weight in Chapter 1, the small-raindrop terminal
velocity in Chapter 2, the recalibration interval for a kitchen scale
in Chapter 3, the pendulum $g$-measurement uncertainty in Chapter 4,
and the analog-to-digital converter resolution required for the
thermistor measurement in Chapter 5. The environment has been used
informally; this section codifies its formal structure.

\subsection{The four moves}

A formal estimation block contains four moves, in the order given.

\engterm{Move 1: question}. The question is stated explicitly, in
bold, with the units of the answer identified. A question whose
units are unclear is not yet a quantitative question.

\engterm{Move 2: record the estimate}. The estimate is performed
and written down, with its decomposition and its assumed factor
ranges, before any calculation or lookup. The reader who omits this
step has not yet entered the estimation discipline.

\engterm{Move 3: compare to the result}. The estimate is compared
to a calculated, measured, or published result. The comparison is
quantitative: the ratio of the estimate to the result is computed,
and the result is recorded as ``within a factor of $k$,'' where
$k$ is the smaller of the ratio and its inverse.

\engterm{Move 4: postmortem}. The factor in the decomposition that
contributed the largest share of the discrepancy is identified.
The working knowledge that would tighten the estimate is named.
The reader records the lesson; the next estimate proceeds from a
tighter anchor.

The four moves are required. A block earns the name \verb|estimation|
only when it records the estimate before lookup, calculation,
measurement, or simulation. A block without a postmortem leaves the
dominant error source unnamed.

\subsection{The blocks already encountered}

The 747 estimate in Chapter 1 used the four-move structure: the
question (mass at maximum take-off weight, in tonnes); the recorded
estimate ($390$ to $430$ tonnes, decomposed into passenger load,
operating empty mass, and fuel); the comparison to Boeing's
$396{,}890 \,\si{\kilogram}$ maximum take-off weight for the 747-400
\cite{web:boeing-commercial-reference}; and the postmortem (the
empty-mass factor was the riskiest move; a beginner's intuition can
mistakenly anchor it on the payload).

The raindrop estimate in chapter 2 used dimensional analysis to
construct the decomposition itself. The decomposition was the
balance between gravity and inertial drag; the factors were the
drop diameter, the drop and air densities, and gravity. The
postmortem identified the inertial-drag-dominated regime as the
working approximation.

The kitchen-scale estimate in chapter 3 was a deep-uncertainty
problem of the kind described in section 9.5. The question was a
recalibration interval; the answer was a range bounded by the
user's tolerance, the scale's repeatability, and the absence of a
manufacturer drift specification. The postmortem identified the
headroom-to-tolerance ratio as the lever the user could pull.

The pendulum $g$-uncertainty estimate in chapter 4 was a
sensitivity analysis: which input dominates the output's
uncertainty. The decomposition was the linear propagation rule;
the factors were the input relative uncertainties weighted by
their sensitivity coefficients. The postmortem identified the
time measurement, with its squared exponent, as the dominant
contributor.

The thermistor ADC-resolution estimate in chapter 5 was an
instrument-specification estimate. The decomposition was the
voltage-divider relation; the factors were the supply voltage,
the divider ratio, and the ADC bit count. The postmortem
identified the temperature-dependent divider sensitivity as the
factor that changes with operating point.

Each of the five blocks performed the four moves, with the
postmortem turning the comparison into a learning event. This
chapter's section 9.1 added a sixth: the Chicago piano-tuner
estimate, which is the canonical Fermi-method demonstration.

\subsection{Where the environment recurs}

The environment recurs in every chapter of every later volume. The
reader who finishes the book has performed estimates in chapters on
mathematics, physics, chemistry, biology, information, machines,
systems, failure, design, and civilization. The cumulative count
across the twelve volumes is in the high hundreds; the reader who
finishes has the internal ruler the discipline aims to install.

The environment is also the formal home of two adjacent practices.
\engterm{Estimate-before-measure} is the same discipline applied to
a measurement that has not yet been taken; the reader records the
estimate before reading the instrument. \engterm{Estimate-before-
simulate} is the same discipline applied to a simulation that has
not yet been run; the reader records the estimate before the
simulator's first output. Both practices defend against the same
confirmation bias the estimate-before-calculate discipline
defends against, and both use the same four moves.

\subsection{Calibration: the reader's own track record}

The fifty Fermi problems at the end of the chapter are also the
reader's first calibration dataset. A reader who records, for each
estimate, both a point value and a stated $90\,\%$ confidence
interval can later compute the empirical hit rate of the intervals
against the published ground truth. A well-calibrated reader's $90$
percent intervals contain truth on $90$ percent of the problems;
\Cref{fig:vol01:ch09:calibration-plot} shows the typical
overconfident shape that an untrained estimator's first calibration
plot takes, and the diagonal it can be pulled toward by repeated
practice. The Brier score is the standard scalar summary of
calibration quality \cite{paper:ch09-brier-1950}; a reference
implementation is in
\repopath{volumes/01-quantity/ch09-discipline-of-estimation/code/brier_score.py}.

\input{volumes/01-quantity/ch09-discipline-of-estimation/figures/fig-calibration-plot.tex}

\begin{mastery}
The Good Judgment Project, an open forecasting tournament run from
$2011$ to $2015$, demonstrated that calibration is a trainable skill
\cite{text:ch09-tetlock-superforecasting-2015}. A small fraction of
participants (the \emph{superforecasters}) consistently beat the
aggregate by $30\,\%$ to $40\,\%$ on the project's Brier score. The
techniques the superforecasters used overlap heavily with the four
moves of this chapter: explicit decomposition, recorded
intermediate estimates, regular postmortems against ground truth,
and a habit of revising forecasts when new evidence arrives. The
implication for engineering practice is direct: the reader who
treats the fifty Fermi problems as the first month of a
multi-year calibration practice acquires forecasting skill that
transfers to project planning, reliability prediction, and the
risk-rate estimates this chapter's failure section is about.
Cooke's expert-judgement method, which weights experts by their
calibration performance on seed questions whose answers are known,
extends the same discipline to risk assessment in industries from
nuclear power to aviation \cite{paper:ch09-cooke-expert-2008}.
\end{mastery}

The same statistical mechanism that makes calibration work for the
individual also makes it work for crowds.
\Cref{fig:vol01:ch09:crowd-convergence} plots the running geometric
mean of independent estimates against the truth, reproducing the
$1906$ Galton ox-weight result: individual estimates scatter, the
geometric mean converges to truth as $1/\sqrt{n}$. The engineering
implication is that several independent decompositions of the same
problem, aggregated in log space, are tighter than any one of them.
A reader who has time for only one Fermi estimate accepts the
factor-of-three working target; a reader who has time for five and
aggregates them can reach a factor-of-two target without any one
decomposition becoming more accurate.

\input{volumes/01-quantity/ch09-discipline-of-estimation/figures/fig-crowd-convergence.tex}

\section{Failure: the Challenger O-ring failure rate}\pathtag{core}

A failure-rate estimate can become dangerous when it travels through
an organisation without an independent sanity check. The Space
Shuttle Challenger O-ring case is the chapter's canonical example,
read here through the dissenting appendix Richard Feynman attached to
the Rogers Commission report \cite{acc:nasa-challenger,paper:ch09-feynman-appendix-f}.
The full technical and organisational case study lives in Volume X;
the present treatment is restricted to the estimation discipline
that the case bears on.

\subsection{The case in brief}

The Space Shuttle Challenger broke up $73$ seconds after launch on
$28$ January $1986$. The proximate cause was failure of an O-ring
seal in the right solid rocket booster's aft field joint at an
ambient launch temperature of approximately $-2 \,\si{\degreeCelsius}$.
Hot combustion gas escaped through the joint, damaged the external
tank attachment region, and led to vehicle breakup
\cite{acc:nasa-challenger}. The launch decision occurred after
Morton Thiokol engineers had objected to launch under the expected
cold-temperature conditions. The Rogers Commission report is the
primary source; Diane Vaughan's 1996 analysis is the canonical
secondary source on the organisational mechanism
\cite{hist:vaughan1996}.

The Challenger case is recorded in the named-cases registry at
\repopath{docs/research/accidents/challenger-1986.md}. This chapter's
framing differs from the volumes that follow: Volume V draws the
material lesson (elastomer cold-temperature behaviour); Volume IX
draws the reliability lesson (redundant seals whose failures are
correlated); Volume X draws the organisational lesson (the
normalization of deviance and the launch-decision teleconference);
Volume XI draws the ethics lesson. We draw the estimation lesson.

\paragraph{Technical mechanism.}
The SRB's segmented case (manufactured by Morton Thiokol at the
Promontory, Utah plant) was joined by tang-and-clevis field
joints sealed by two redundant Viton fluorocarbon-elastomer
O-rings (Parker Hannifin part number S0604-70, inner diameter
$2785\,\si{\milli\meter}$, cross-section $7.0\,\si{\milli
\meter}$). At normal operating temperatures
($24$-$30\,\si{\degreeCelsius}$), the rings deformed elastically
under the joint's pressure-driven rotation (the inner case
expands outward by $0.7\,\si{\milli\meter}$ at SRB ignition
pressure of $\sim 6.4\,\si{\mega\pascal}$) and maintained the
seal. The Viton elastomer's glass transition temperature is
approximately $-15\,\si{\degreeCelsius}$; near and below freezing
the elastomer's resilience drops sharply. STS-51-L's launch-time
ambient temperature was $-2.2\,\si{\degreeCelsius}$ ($28\,
{}^{\circ}\mathrm{F}$), the lowest at which the Space Shuttle had
ever launched (next lowest: STS-51-C in January 1985 at
$11.7\,\si{\degreeCelsius}$, on which significant O-ring erosion
had been observed and documented). Computed local O-ring
temperatures from the wind-cooled aft field joint were $\sim
-1.6\,\si{\degreeCelsius}$. The O-ring's response time
(measured Compression Set Recovery) at $-2\,\si{\degreeCelsius}$
was approximately $600\,\si{\milli\second}$, compared to $\sim
50\,\si{\milli\second}$ at $20\,\si{\degreeCelsius}$; the
post-ignition joint rotation occurred over $\sim 600\,\si{
\milli\second}$, the same time scale.

At ignition at $11{:}38{:}00.010$ EST on $28$ January $1986$, the
right SRB's aft field joint failed to seal completely. Black smoke
ejected from the joint over the first $2.5\,\si{\second}$ of
flight (Range Camera E-60 footage frames show $9$ distinct puffs
of smoke, spaced approximately $0.3\,\si{\second}$ apart, from
the aft field joint), then the joint appeared to reseal as
aluminium-oxide combustion products solidified in the gap. At
$T+58.788\,\si{\second}$ wind shear at $32\,\si{\kilo\meter}$
altitude (Mach $1.92$) loosened the temporary plug; a persistent
flame became visible on E-207 camera footage at $T+59\,\si{
\second}$. The flame burned through the lower attachment strut
between the SRB and the external tank by $T+72\,\si{\second}$
and through the aft dome of the external tank by $T+73.2\,\si{
\second}$. The hydrogen tank failed; the orbiter, still under
aerodynamic loads at Mach $1.92$ and dynamic pressure
$33.9\,\si{\kilo\pascal}$, broke up at $T+73.6\,\si{\second}$.
The crew compartment, intact but unpowered, struck the Atlantic
Ocean at $T+170\,\si{\second}$ at approximately $325\,\si{\kilo
\meter\per\hour}$ \cite{acc:nasa-challenger}.

\paragraph{Organisational context.}
Engineers at Morton Thiokol had identified the O-ring's
temperature-dependent resilience as a concern through internal
Thiokol memos dating from August 1985, culminating in a
$23$ July $1985$ memo by Roger M. Boisjoly (Thiokol staff
engineer) to Vice President of Engineering R. K. Lund identifying
the joint as a "criticality 1" failure mode and recommending an
engineering task force, which was constituted but not adequately
resourced. On the evening of $27$ January $1986$, Boisjoly and
fellow Thiokol engineer Arnie Thompson presented charts in a
teleconference with NASA Marshall Space Flight Center showing
that O-ring blow-by had been observed on every flight at launch
temperature below approximately $17\,\si{\degreeCelsius}$, and
recommended against launch outside the qualified envelope of
$11.7$-$32\,\si{\degreeCelsius}$. Thiokol Vice President Joe
Kilminster, after a $30$-minute internal caucus during which the
data presentation was reviewed by Thiokol senior management (Bob
Lund, Jerry Mason, Cal Wiggins, and Joe Kilminster), reversed the
recommendation; the four-person management caucus voted
unanimously to recommend launch despite the engineering
objection. Lund reportedly said to Mason during the caucus, in
testimony later given to the Rogers Commission, "take off your
engineering hat and put on your management hat."

NASA Marshall management (Lawrence Mulloy, SRB project manager;
George Hardy, Marshall deputy director) accepted the reversed
recommendation. The Level III flight readiness review had been
conducted on $25$ January at Cape Kennedy; the temperature
concern was not surfaced at the higher Level II review at
Marshall on $26$ January or at the Level I review at NASA
Headquarters on $27$ January. The reversal teleconference of the
evening of $27$ January was therefore the last decision point
that could have stopped the launch; the next morning, no
additional review preceded the count-down sequence
\cite{acc:nasa-challenger,hist:vaughan1996}. Vaughan's analysis
of the same documents identified the "normalization of deviance",
the pattern in which signals of departure from design
specification (O-ring erosion observed on $S T S$-$2$, $-41B$,
$-41C$, $-41D$, $-51C$, $-61A$, $-61C$) were progressively
reinterpreted as within acceptable risk, rather than as evidence
that the original specification had not been met
\cite{hist:vaughan1996}. Allan J. McDonald, Thiokol's manager at
the Kennedy Space Center who had refused to sign the launch
recommendation, attempted unsuccessfully to halt the launch
through NASA channels; his name and the disposition of his
objection appear in the Rogers Commission report
\cite{acc:nasa-challenger}, but the management chain did not
preserve his veto into the launch decision.

\paragraph{Lesson for estimation discipline.}
The estimation lesson is concentrated in Appendix F of the Rogers
Commission report, the dissenting appendix that Feynman insisted be
included \cite{paper:ch09-feynman-appendix-f}. Feynman's appendix is
the canonical primary source for the chapter's claim about the
discipline, because it documents both the mismatched estimates and
the way they coexisted in the same programme without reconciliation.
The next two subsections work the appendix's evidence through the
chapter's four moves.

\subsection{The unsanity-checked estimate}

Appendix F of the Rogers Commission report records estimates of
loss-of-vehicle-and-crew probability that range from roughly $1$ in
$100$ to $1$ in $100{,}000$ \cite{paper:ch09-feynman-appendix-f}.
Feynman reports that the higher-risk estimates came from working
engineers and the lower-risk estimates from management; he also
summarises range-safety evidence from prior solid rocket flights as
far closer to the $10^{-2}$ scale than to $10^{-5}$. The appendix's
closing line is the one most often quoted: ``For a successful
technology, reality must take precedence over public relations, for
nature cannot be fooled.''

The discrepancy of three orders of magnitude is the discrepancy this
section is about. The lower estimate, $10^{-5}$, was the working
number used in the programme. The higher estimate, $10^{-2}$, was an
independent check. By the discipline this chapter has been training,
a discrepancy of three orders of magnitude between two estimates of
the same quantity is a release-blocking signal: the engineer cannot
proceed without investigating which estimate is wrong, or whether
both are wrong in different directions.

The discrepancy did not receive the level of investigation it
demanded. A top-down management number near $10^{-5}$ and engineering
estimates near $10^{-2}$ cannot both guide the same launch decision
without reconciliation. Retrospectively, the Shuttle programme
suffered two catastrophic losses in $135$ flights, Challenger in 1986
and Columbia in 2003, a rate of about $1.5 \times 10^{-2}$ over the
completed programme \cite{acc:nasa-challenger,acc:caib-columbia}.
That retrospective figure does not prove what could have been known
before STS-51-L, but it shows which order of magnitude the historical
programme eventually occupied. The engineering estimates of 1985 were
within a factor of two of the eventual programme record; the
management estimate was wrong by three orders of magnitude.

\subsection{What the discipline would have caught}

The estimation discipline, applied at the level we have trained in
this chapter, would have stopped at the three-orders-of-magnitude
discrepancy. The discipline would have asked: which of the two
estimates is bottom-up, derived from component failure data and the
algebra of redundant-system reliability; which is top-down, derived
from a budgetary or programmatic allocation; which estimate is
consistent with the historical record of comparable systems; which
estimate is consistent with the observations of erosion and blow-by
in prior flights' O-ring post-flight inspections.

Each of those questions points away from $10^{-5}$. Feynman's
Appendix F reports working estimates around $1$ in $100$ and
engineering estimates for related Shuttle subsystems as high as $1$
in $300$ or $1$ to $2$ per $100$, while management still claimed $1$
in $100{,}000$ \cite{paper:ch09-feynman-appendix-f}. The observations
of O-ring erosion and blow-by on prior flights, treated as data
rather than as accepted anomalies, also push the estimate toward the
higher-risk order of magnitude. The lower number was the outlier; the
discipline would have required an explanation before using it.

Feynman's worked example in the appendix is itself a model Fermi
estimate. He decomposes the launch-failure probability through the
known number of solid-rocket failures in similar programmes
(approximately one in twenty-five for U.S.\ unmanned launches at the
time), discounts for the Shuttle's additional safety review, and
arrives at a working estimate of about one in a hundred per flight.
The decomposition is short, every factor is defensible from primary
data, and the result is within a factor of two of the eventual
two-loss-in-$135$-flight record. The estimate is what the chapter has
been training: question stated, factors named, decomposition
recorded, postmortem against the historical record built in. The
appendix is the chapter's case study in why the four moves matter.

The Vaughan analysis identifies, at the organisational level, the
mechanism by which the outlier was preserved \cite{hist:vaughan1996}.
The estimation discipline of this chapter is one defence against the
pattern. An engineer who has performed the bottom-up reliability
calculation, recorded the result, and compared it to the
historical-record estimate cannot, without effort, then adopt a
budgetary number that disagrees with both by three orders of
magnitude. The discipline is also a defence at the report level: an
estimate that travels with its decomposition and its
historical-record check is harder to overwrite by a number from
elsewhere in the organisation.

\subsection{A 1985-era quantitative pre-launch risk estimate}

The chapter's framing of the Challenger case turns on whether a
defensible quantitative estimate of the O-ring failure rate could
have been constructed from the data available before STS-51-L. The
short answer is yes; the longer answer is the worked example of
this subsection, which recovers what a $1985$-era engineer could
have computed and contrasts the result with the management estimate
the launch decision used.

\paragraph{Step 1: the data the engineer had.}
By January $1986$, the Shuttle programme had completed $24$ flights.
Of those, $7$ had returned with observable O-ring damage on at least
one field joint: erosion of the primary O-ring, evidence of hot-gas
blow-by, or both. The damage events were recorded in post-flight
inspection reports, and the ambient temperatures at launch for each
flight were also recorded. The combined dataset is reproduced in
Appendix F of the Rogers Commission report and was reanalysed
post-disaster by Dalal, Fowlkes, and Hoadley
\cite{paper:ch09-feynman-appendix-f,
ch09-paper:dalal-challenger-1989}.

\paragraph{Step 2: a frequentist point estimate.}
The simplest defensible estimate treats the $7$ damaged flights as
realisations of a Bernoulli process with success probability $p$
(``success'' = ``damage observed''). The maximum-likelihood point
estimate is $\hat p = 7/24 \approx 0.29$. The Wilson $95\,\%$
confidence interval for a binomial proportion with $7$ events in
$24$ trials is approximately $[0.14, 0.50]$. Even setting aside the
temperature dependence, the empirical damage rate was already an
order of magnitude per flight $10^{-1}$, with damage events
occurring at roughly one flight in three; the management estimate
of $10^{-5}$ disagreed with the bare empirical rate by four orders
of magnitude.

\paragraph{Step 3: the temperature signal.}
The $7$ damaged flights had a different temperature distribution
from the $17$ undamaged flights. The dataset's bare temperatures at
launch ranged from approximately $12 \,\si{\degreeCelsius}$ to
$27 \,\si{\degreeCelsius}$ for the historical record. A logistic
regression of damage outcome against launch temperature gives a
fitted relation
\[
\Pr(\text{damage} \mid T) = \frac{1}{1 + \exp(\beta_{0} + \beta_{1}
T)},
\]
with parameters estimated post-hoc by Dalal et al.\ at $\beta_{0}
\approx -5.1$ and $\beta_{1} \approx 0.12 \,\si{\per\degreeCelsius}$
on the published data \cite{ch09-paper:dalal-challenger-1989}. The
fit is statistically significant against the null of
temperature-independence at $p < 0.05$ on the pre-Challenger
record.

\paragraph{Step 4: extrapolate to the launch temperature.}
The launch temperature on $28$ January $1986$ was approximately
$-2 \,\si{\degreeCelsius}$, well below the minimum launch
temperature in the prior dataset. Extrapolating the logistic
relation to this temperature gives $\Pr(\text{damage} \mid T =
-2) = 1 / (1 + \exp(-5.1 + 0.12 \times -2)) = 1 / (1 + \exp(-5.34))
\approx 0.995$. The estimate's confidence interval is wide because
the extrapolation is well outside the calibrated range; a defensible
range is $[0.70, 1.00]$, but the point estimate sits near unity.
Even the lower bound of $0.70$ is many orders of magnitude above
$10^{-5}$.

\paragraph{Step 5: combine the damage probability with the
failure-given-damage probability.}
A damaged O-ring is not by itself a loss of vehicle. The secondary
O-ring exists precisely to recover from primary erosion, and prior
flights with damage had returned successfully. A defensible
estimate of $\Pr(\text{loss of vehicle} \mid \text{damage})$ depends
on whether the primary and secondary O-rings fail independently or
in correlated fashion. The pre-Challenger engineering analysis
assumed they failed independently: $\Pr(\text{both fail} \mid
\text{either fails}) \approx \Pr(\text{secondary fails} \mid
\text{primary fails}) \approx 0.1$, giving a per-flight loss
probability of $\Pr(\text{damage}) \times 0.1 \approx 0.03$ at the
mean historical temperature, or $\approx 0.07$ at the
$-2 \,\si{\degreeCelsius}$ extrapolated temperature, with the
caveat that the independence assumption is unsupported at low
temperature where the same elastomer property degrades both seals
\cite{ch09-paper:dalal-challenger-1989}.

\paragraph{Step 6: contrast with the management estimate.}
The bottom-up estimate, even with conservative independence
assumptions, lands in the range $10^{-2}$ to $10^{-1}$ per flight
at the historical mean temperature, and substantially higher at
the extrapolated launch temperature. The management number used in
programme-level risk budgeting was $10^{-5}$. The gap is between
three and four orders of magnitude. By the chapter's principle, an
estimate of $10^{-5}$ cannot coexist with a bottom-up estimate of
$10^{-2}$ without one of the two being explicitly retracted; an
organisation that runs the launch decision against the $10^{-5}$
figure without resolving the discrepancy has not performed the
sanity check the discipline requires.

The post-Challenger record made the calculation visible. Two
catastrophic losses in $135$ flights gives a programme loss rate of
$\approx 1.5 \times 10^{-2}$ per flight
\cite{acc:nasa-challenger,acc:caib-columbia}. The pre-launch
quantitative estimate available from the $1985$ data, with the
independence assumption that the analysis used at the time, lands
within a factor of two of the eventual record. The estimate was
not missing for lack of data; it was missing for lack of the
discipline that would have caused the data to be aggregated and
compared to the management number before the launch decision.

\subsection{The qualitative pre-launch reasoning and what
quantification would have shown}

The $1985$-era engineers at Thiokol who objected to the launch
had a defensible qualitative argument: O-ring blow-by had been
observed on every flight whose joint temperature was below about
$17 \,\si{\degreeCelsius}$, and the $28$ January launch
temperature was many degrees below the coldest prior data point.
The defensible engineering position was that the launch was
outside the qualified envelope. That position was the recorded
view at the start of the $27$ January teleconference.

The objection's working strength was bounded by what the
Thiokol engineers had quantified and what they had not. Boisjoly's
$23$ July $1985$ memo to R.~K.~Lund \cite{paper:ch09t3-boisjoly-1985}
identified the joint as a criticality-1 failure mode and
recommended an engineering task force; the memo did not contain
a numerical estimate of the failure probability under cold
launch conditions, because the qualification data set did not
extend to those conditions. The $27$ January teleconference
slides plotted the prior O-ring damage events against
temperature; the plot was qualitative, showing that low-
temperature flights had damage and warm-temperature flights had
less, but it did not include flights without damage (a known
selection bias the post-Challenger statistical reanalysis
corrected for; see Step 3 above) and it did not produce a
probability number for the launch conditions. The objection was
defensible at the qualified-envelope level; it was not defensible
at the level of a quantified probability estimate.

A $1985$-era quantitative estimate, performed on the data the
engineers already had, would have produced exactly the
calculation that section~9.8.4 walked through. The logistic
regression of damage against temperature, performed in $1985$
on the $24$ pre-Challenger flights, produces the same parameters
that Dalal, Fowlkes, and Hoadley recovered post-disaster
\cite{ch09-paper:dalal-challenger-1989}; the inputs were available
before the launch, only the calculation was missing. The
extrapolation to the launch temperature, even with its wide
confidence interval, would have placed the $P(\text{damage})$
near $1$ rather than at the historical mean of $\approx 0.3$. A
Bayesian update, performed on a prior centred on the historical
$\approx 0.3$ damage rate and weighted by the temperature
evidence, would have shifted the posterior for $P(\text{seal
failure})$ at the launch conditions from approximately $0.001$
(no-blow-by under nominal conditions) to approximately $0.05$
(blow-by likely at $-2 \,\si{\degreeCelsius}$). Either
calculation, presented at the $27$ January teleconference, would
have shifted the working number that the launch decision had
to refute from a vague ``outside the qualified envelope'' to a
specific ``one-in-twenty probability of seal failure on this
launch.'' The vague form was easier for management to override
than the specific form would have been.

The estimation lesson is the inverse of the LTCM lesson. Where
LTCM had a model and not enough data, Challenger had data and
not enough model. The available data, treated as data and
quantified, would have produced a probability number that
management could not have dismissed without recording a
specific refutation. The qualitative objection, by contrast,
was overridden in the teleconference without a recorded
refutation, because there was no specific number to refute.
The Fermi discipline at the level of section~9.8.4 supplies
the specificity that the qualitative objection lacked. The
record discipline at the level of the principle below would
have required either the quantification or a documented
counter-argument before the launch decision could proceed.

\subsection{The post-Challenger shift to evidence-based
engineering}

The Rogers Commission's report, issued $6$ June $1986$,
recommended nine specific actions, of which Recommendation~III
(``the inherent flaws of the SRB joint and seal must be
fully understood'') drove a $\$1.3$~billion Solid Rocket Motor
Redesign program at Thiokol over the years $1986$ through
$1988$ \cite{paper:ch09t3-rogers-implementation-1987}. NASA's
formal Implementation Plan, issued in July $1987$, also
reorganised the SRM Project Office at Marshall Space Flight
Center, established a Headquarters-level Office of Safety,
Reliability, Maintainability, and Quality Assurance, and
formalised the criticality-1 flight-readiness-review process
that the pre-Challenger reviews had not enforced
\cite{paper:ch09t3-rogers-implementation-1987}.

The deeper procedural shift the Implementation Plan formalised
was the inversion of the burden of proof. Before Challenger,
the working procedural rule had been that an anomaly observed
in flight could be flown again unless the engineering data
proved it would fail (``prove it WILL fail''). The Rogers
Commission recommendations and the post-Challenger procedural
manuals formalised the inverse: an anomaly observed in flight
holds the next flight until the engineering data prove it
will not fail (``prove it WON'T fail''). The inversion is the
post-Challenger SRM redesign's most consequential procedural
output, larger than any specific technical change to the
joint, because it shifted the default state of an anomaly
from ``proceed'' to ``hold.''

Vaughan's later analysis named the pre-Challenger pattern the
\emph{normalization of deviance}: a sequence of in-flight
anomalies, each of which had been individually classified as
within acceptable risk because the prior anomalies had not led
to loss, gradually moved the working risk perception away from
the engineering data \cite{hist:vaughan1996}. The
burden-of-proof inversion is the procedural defence against
the pattern. Under the post-Challenger rule, the cumulative
evidence of O-ring blow-by on $7$ of $24$ prior flights would
have held the launch pending a defensible re-quantification of
the seal failure rate; under the pre-Challenger rule, the
same evidence had been reinterpreted as within acceptable
risk because no prior flight with blow-by had been lost. The
estimation discipline is the procedural defence at the
working-engineer level; the burden-of-proof rule is the same
defence at the institutional level. The two reinforce each
other when both are in place and reinforce each other's
failure when neither is.

McDonald's first-hand account \cite{hist:ch09t3-mcdonald-2009}
records the burden-of-proof inversion from the Thiokol-Kennedy
side. McDonald was the on-site Thiokol manager who refused to
sign the launch recommendation; his refusal was overridden
under the pre-Challenger procedural rule, because the
management chain accepted the reversed Thiokol senior-
management recommendation as the ``proves it will fail''
threshold being unmet. The post-Challenger rule, applied to the
same teleconference inputs, would have required the management
chain to demonstrate that the cold-temperature blow-by
mechanism could be ruled out before the launch could proceed.
The same evidence, the same engineers, and the same management
chain produce opposite launch decisions under the two rules.
The procedural shift is the lesson Vol I Ch 9 carries forward;
the technical and material aspects of the redesign appear in
Vols V, IX, and X.

The estimation discipline's role in the burden-of-proof
inversion is the supply of the specific number against which
the rule operates. ``Prove it won't fail'' is an empty rule
unless there is a specific failure probability against which to
test the proof; the engineering action under the rule is to
produce the quantitative estimate that the management decision
must then either accept or refute. The chapter's four moves
are the working machinery that produces the number. The
pre-Challenger absence of the number was the procedural gap
that the post-Challenger rule exists to fill.

\subsection{The general lesson}

A defensible failure-rate estimate is checked against component
data, the historical record of comparable systems, and observed
in-service anomalies. The Challenger case is the canonical instance
because the disagreement between the management estimate and the
available cross-checks was three orders of magnitude. The same
failure mode can appear whenever a programme adopts a reliability
number for budgetary or scheduling purposes without the calculation
the number appears to represent.

The principle is the chapter's closing rule.

\begin{principle}[Estimates are sanity-checked against independent
estimates, or they do not leave the desk]
A reported estimate of a quantity that matters is sanity-checked
against at least one independent estimate, derived by a different
method or from different data. When the two agree within their
stated uncertainty, the estimate is reportable. When they
disagree by more than the stated uncertainty, the engineering
action is to investigate the disagreement before reporting any
estimate at all. An estimate that has not been cross-checked
against an independent route is, by this discipline, a working
number that is not yet ready to leave the desk.
\end{principle}

\section{Historical sketch: from Fermi's piano tuners to
Mahajan's pedagogy}\pathtag{standard}

The order-of-magnitude habit predates Enrico Fermi by a
century or more (Lord Rayleigh's $1894$ dimensional-analysis
discussion in \emph{The Theory of Sound} is a working
ancestor; Lord Kelvin's $19$th-century estimate of the age of
the Earth is another), but the genre carries Fermi's name
because his teaching practice at the University of Chicago in
the $1940$s formalised it and because his working physics
practice put it to consequential use.

\subsection{Fermi's pedagogy and the Trinity estimate}

Fermi had moved from Columbia University to the University of
Chicago's Metallurgical Laboratory in $1942$, where he led the
team that achieved the first sustained nuclear chain reaction
in CP-1 on $2$ December $1942$. His Chicago physics seminars
of the late $1940$s posed quantitative questions whose answers
were not in any reference: how many piano tuners work in
Chicago, how many atoms in a breath of air, how long for a
single drop of water to evaporate from a swimming pool in
summer. The pedagogy was deliberate. A graduate student who
could decompose an unfamiliar question into factors each
defensible from working knowledge had the working apparatus
of a physicist; a student who could only solve textbook
problems did not.

The pedagogy's most consequential public application was
Fermi's own paper-strip estimate of the Trinity test yield on
$16$ July $1945$. Fermi stood approximately $9.6 \,\si{\kilo
\meter}$ from the detonation site, holding a stack of small
paper strips. He released the strips at the moment of the
shock-wave arrival and measured the horizontal displacement
the strips experienced as the shock passed: approximately $2.5
\,\si{\meter}$. From the displacement, the distance to the
source, and a working approximation that the shock wave's
kinetic energy was a fixed fraction of the detonation yield,
Fermi inferred a yield of approximately $10 \,\text{kT}$ of
TNT equivalent on the spot \cite{hist:ch09t3-fermi-trinity}.
The eventual calorimetric reconstruction, performed by
detailed instrumentation and reported weeks later, gave the
yield as approximately $21 \,\text{kT}$
\cite{hist:ch09t3-rhodes-1986}. Fermi's paper-strip estimate
was within a factor of two of the eventual answer, derived in
seconds using equipment that would not have impressed a high-
school physics teacher.

The Trinity estimate is the historical anchor of the
chapter's discipline. The estimate's economic case (no
instrumentation cost) was unbeatable; its precision
(factor-of-two on a quantity that had never been measured
before) was sufficient for its working use (telling the
observers what order of magnitude they had just witnessed);
and its postmortem (the eventual calorimetric reconstruction)
identified the dominant error source (the shock-wave-energy-
to-total-yield ratio that Fermi had approximated as $1/2$
where the true ratio was closer to $1/3$). The four moves the
chapter has been training are present in the
Trinity estimate. Fermi's paper-strip work is the genre's
founding case.

\subsection{The mid-century period and the popular tradition}

Between the $1940$s and the $1980$s the order-of-magnitude
habit lived inside specific working communities (physicists,
operations researchers, nuclear engineers) without a popular
text. Three working factors changed that during the $1980$s.
The personal computer brought spreadsheets into engineering
offices, and the spreadsheet's silent-error mode (units,
sign, decimal place) made the estimate-before-calculate
defence visible to a wider working population. The
back-of-envelope tradition in operations research, formalised
in textbooks such as Wagner's $1969$ \emph{Principles of
Operations Research} and updated through the $1970$s,
established the discipline at the level of supply chains and
queueing systems. And John Allen Paulos's $1988$ \emph{Innumeracy}
\cite{text:ch09t3-paulos-1988} brought the order-of-magnitude
habit into public discourse, with chapter-length working
examples on the size of large numbers, the probability of
coincidences, and the working precision of mortality and
risk statistics.

Paulos's contribution to the chapter's discipline is the
public-discourse argument: a citizenry that cannot perform
order-of-magnitude estimates cannot evaluate the public-policy
claims that depend on them. The argument is not unique to
$1988$ (Bertrand Russell had made a version of it in $1931$,
and Carl Sagan had made another in $1973$), but Paulos's
specific framing of the working examples, with the
arithmetic written out, is the version that entered the
engineering and the science-communication canon. The
chapter's discipline is the working-engineer version of the
same defence; the order-of-magnitude habit that protects an
engineer's calculation from a unit error is the same habit
that protects a citizen's policy judgement from a
decimal-point error.

\subsection{Mahajan and the contemporary pedagogy}

The pedagogy's contemporary form crystallised at MIT in the
$2000$s through the working of Sanjoy Mahajan. Mahajan's
$2008$ MIT course \emph{6.055J / 2.038J: The Art of
Approximation in Science and Engineering}, recorded on MIT
OpenCourseWare \cite{web:ch09t3-mit-6055j}, codified the
working moves into a teachable sequence: divide-and-conquer
decomposition, abstraction, dimensional homogeneity, easy
cases, log-scale uncertainty, and the postmortem against
ground truth. The course became the $2010$ MIT Press book
\emph{Street-Fighting Mathematics}
\cite{text:ch09t3-mahajan-streetfighting-2010} and the $2014$
sequel \emph{The Art of Insight in Science and Engineering}
\cite{text:mahajan2014}, both released under Creative Commons
licences and available without charge in PDF.

Mahajan's contribution is the structured teaching of the
discipline. The Fermi pedagogy had been a teacher's private
practice; the Paulos popularisation had been a public
argument; Mahajan's books are the working pedagogy that an
instructor can hand to a student. The chapter's structure (the
four moves, the log-quadrature combination rule, the
postmortem requirement, the calibration practice) tracks the
structure Mahajan codified, with engineering examples
substituted for the physics examples Mahajan worked through.
The reader who finishes the chapter and wants more practice
can take Mahajan's MIT course directly; the course is the
contemporary canonical reference for the discipline at the
undergraduate level.

The pedagogy has continued to spread. The Stanford ME~$203$
course on engineering estimation, taught from $2012$, uses
Mahajan's framework. The Caltech APh~$5/105$ course on the
physics of nanoscale systems uses Fermi-style estimation as
its working homework genre. The German engineering education
reform of the $2010$s introduced \emph{\"Uberschlagsrechnung}
(literally ``rough-pass calculation'') as a required
competency in the \emph{Diplom-Ingenieur} curriculum. The
chapter's discipline is now a working part of the engineering
canon at the institutions that train the engineers the book
intends to serve.

\subsection{Why orders of magnitude became part of the
engineering canon}

The discipline's path into the engineering canon has three
working drivers. The first is the economic case from the
$1980$s onward: the cost of a missed estimate at detailed
design is two to three orders of magnitude higher than the
cost of the estimate at pre-feasibility. The asymmetric
cost-of-rework is what makes the design habit of
section~9.5 a working discipline rather than a teaching
exercise. The Boeing $787$ case is the canonical recent
instance; comparable cases run through the late twentieth
and early twenty-first century engineering record.

The second driver is the spreadsheet's silent-error mode.
The spreadsheet computes whatever the formula says, and the
formula can have a unit error, a sign error, or a
decimal-point error that the displayed number reveals only
to a reader who has an independent expectation. The
estimate-before-calculate discipline supplies the
independent expectation. The discipline became a working
part of the engineering canon at roughly the same time the
spreadsheet became the working calculation tool, because
the same factor of $10^{3}$ or $10^{6}$ that a silent unit
error introduces is the same factor that the order-of-
magnitude estimate catches.

The third driver is the asymmetry between calculation and
judgement. A calculation that produces a number is
sometimes mistaken for a judgement; a judgement that
produces a number is sometimes mistaken for a calculation.
The estimation discipline forces the distinction: an
estimate is recorded with its decomposition, its factor
ranges, and its postmortem; a calculation is recorded with
its formula, its inputs, and its result; the comparison
between them is recorded with the ratio and the
investigation of any disagreement. The discipline keeps
the two categories separate by documenting each
appropriately. The engineering canon adopted the discipline
because the alternative, conflating calculation and
judgement, produces the silent failure modes that the
chapter's accident cases illustrate.

The reader at the close of Volume I is reading the
discipline as the contemporary engineering canon now
teaches it: as the working habit by which a quantitative
question is decomposed, estimated, calculated, compared,
and postmortemed. The genre's founder was Fermi; its
popular advocate was Paulos; its contemporary pedagogue is
Mahajan; its working practitioners are the engineers who
build the reader's bridges, drugs, code, vehicles, and
power grids. The chapter's project at section~9.10 asks the
reader to join the working community.

\section{Estimation as a moral discipline}\pathtag{core}

We close Volume I where the volume began: with the claim that
measurement and engineering are the same discipline at different
scales. Estimation is the form measurement takes when the
instrument is not yet at hand, when the experiment has not yet
been performed, when the calculation has not yet been written.
The reader who carries the discipline forward into Volume II and
beyond carries with them the habit by which a number, even when
it is approximate, is honest.

The discipline has professional content. An engineer who records
an estimate before calculation is harder to move toward a number
selected for institutional convenience. The piano-tuner estimate
is a small case; the Challenger O-ring failure-rate estimate is a
large one. The same record discipline applies to both. An engineer
who declines to report a failure rate derived from allocation
rather than from component data preserves the distinction between
what is known, what is assumed, and what remains to be checked.

The discipline scales further still because records accumulate into
culture. A profession that estimates honestly leaves a traceable
account of its numbers: which came from measurements, which came
from models, which came from judgement, and which were later
checked against reality. A profession that estimates dishonestly
loses that trace. The difference is built one estimate at a time,
repeated across projects, reviews, and years. The reader who
finishes Volume I with the habit installed has done the hardest
part of becoming an engineer in the sense that this work uses the
word.

A working test of the moral content of the discipline is the
Brier score of the reader's calibration record across the volume's
closing fifty Fermi problems
\cite{paper:ch09-brier-1950,text:ch09-tetlock-superforecasting-2015}.
The Brier score of a forecast probability $p$ against a binary
outcome $o \in \{0,1\}$ is $(p - o)^{2}$, averaged across the
forecasts. A constant-prior forecast of $0.5$ on every binary
outcome scores $0.25$; a perfectly calibrated forecaster who
hedges only as much as the underlying ignorance requires scores
better than that; a forecaster who consistently overstates
confidence scores worse. The reader who scores worse than $0.25$
is reporting more confidence than the evidence supports. The
score itself is the discipline's empirical receipt; the reader
who tracks it across projects converts moral discipline into
measurable behaviour.

The Brier score also exposes a second pathology. A forecaster
whose calibration is poor in one direction (systematic over- or
under-confidence) but whose discrimination is high (estimates
above $0.5$ are usually right and estimates below $0.5$ are
usually wrong) carries useful information that a single number
hides. The decomposition of the Brier score into reliability,
resolution, and uncertainty terms is the standard apparatus by
which the working calibrator separates the two failure modes
and trains them out separately
\cite{ch09-paper:bedford-cooke-2001}. The reader is not asked to
master the decomposition here; the reader is asked to record
forecasts in a form that permits the decomposition later. Even a
plain ledger of (forecast, outcome) pairs preserves the
information the discipline needs.

The remaining eleven volumes will build on the habit. Volume II
adds the mathematics that turns each estimated quantity into a
function, a vector, a probability, an algorithm; the estimates
acquire a richer apparatus and a sharper voice. Volumes III
through VIII apply the apparatus to the physical, chemical,
biological, informational, mechanical, and infrastructural
domains. Volumes IX through XII study what happens when the
domains meet: systems, failures, designs, and the civilization
the engineered systems compose. Estimation appears in every one
of those settings, and in each its four moves are the same: ask
the question, record the estimate, compare to the result,
identify the factor that dominated. The reader who has
practised the four moves on fifty Fermi problems at the close
of this volume can practise them, with the same confidence, on
any quantity the rest of the work asks them to estimate.

One small recurring observation is worth recording at this point,
because it returns in Volume II as a mathematical fact and in
Volume IX as a reliability principle. The wisdom-of-crowds effect
in Fermi-problem aggregation, where the geometric mean of $N$
independent estimates often outperforms the median individual
estimate, is itself a consequence of the central-limit theorem
applied to the log-ratios of estimate to truth
\cite{paper:ch09-cooke-expert-2008,
text:ch09-tetlock-superforecasting-2015}. The variance of the
sample-mean log-ratio decreases as $1/N$ in $N$ independent
estimates, so the geometric mean of $N = 25$ independent estimates
has standard error of the log-ratio cut by a factor of five
relative to any single estimator. The empirical evidence on
calibrated forecasters is that the effect saturates around $N \sim
20$ to $40$ when the estimators share working domains, because
the assumption of independence then degrades. The lesson for an
engineer working alone is to estimate several times by
genuinely different routes (bottom-up, top-down, by analogy to a
similar past system, by a conservation argument); the aggregation
of three or four genuinely different routes catches errors that
no single route can.

The fifty Fermi problems below are the volume's closing project.

\subsection{The half-life of the estimate one carries}

Volume I's chapter metadata carries a \engterm{half-life} tag on
every chapter (Q43). The chapter on units carries
``foundational'' because the kilogram and the metre persist
beyond a working career; the chapter on statistics carries
``foundational'' for the same reason; the chapter on data
acquisition carries ``years'' because the specific transducers
shift with each generation. Estimation's half-life is
foundational because the discipline outlives any specific
estimate it produces.

The estimate one carries forward as a working anchor is
typically not the most precise estimate one has performed; it
is the estimate one has performed often enough that the
underlying decomposition has become automatic. The atmospheric
mass of $5 \times 10^{18} \,\si{\kilo\gram}$ is one such
anchor; the body's $\approx 100 \,\si{\watt}$ resting metabolic
power is another; the order-of-magnitude breakdown of energy
sources for a developed country (transport, heating,
electricity, industrial process heat, agriculture: each
contributing $\sim 20\%$ to a working approximation) is a
third. The discipline's payoff is that the anchors accumulate
without bookkeeping; a working engineer who has performed
fifty Fermi estimates carries an internal ruler whose
calibration was free of any conscious effort to maintain it.

The half-life of a specific estimate is the time over which
the underlying inputs change by more than the estimate's
stated range. A population estimate has a half-life of about a
decade; an energy-intensity estimate of about five years; a
data-centre power estimate of about two years; a cryptocurrency
electricity-consumption estimate of less than a year. The
discipline of dating estimates is what permits the reader to
identify the estimate's stale fraction as the years pass. An
estimate dated $2024$ that the reader is reading in $2030$ has
crossed at least one update boundary in many of its inputs;
the working response is not to discard the estimate but to
re-perform the four moves with the updated inputs. The
discipline persists; the specific numbers do not.

\subsection{The engineer's obligation to estimate before
computing}

The chapter's closing claim is a procedural one. An engineer
who computes before estimating produces a number that has not
been checked against an independent expectation, even when the
engineer is in possession of the working knowledge that would
have supplied one. The procedural failure is the silent error
mode the chapter has documented across the Challenger case,
the Boeing $787$ case, the Fukushima case, the LTCM case, and
the climate-stressed-design-load case. In each case the
discipline of recording an estimate before performing the
calculation would have produced a record against which the
calculation's result could have been checked. In each case the
absence of the record permitted the calculation's result to
travel into a decision that the estimate, had it been
recorded, would have blocked.

The procedural rule generalises across domains. The clinician
who calculates a drug dose without an independent
expectation of its order of magnitude permits a unit-error
overdose to be ordered. The pilot who programs a waypoint into
the flight management system without an independent
expectation of its position permits a geographic-coordinate
error to be entered. The financial analyst who runs a Monte
Carlo simulation without an independent expectation of the
output's order of magnitude permits a model-parameter error to
be reported. The procedural defence in each domain is the
same: record an estimate before the calculation, compare
against the calculation when it is done, investigate any
disagreement before reporting.

The Fermi discipline is the engineer's contribution to the
procedural defence. The discipline travels well outside
engineering. Clinicians, pilots, analysts, journalists, and
policymakers each have a working version of the
order-of-magnitude habit; what they share with the engineer is
the procedural rule the chapter has been training. The reader
who has finished Volume I with the habit installed has
acquired a working defence against the silent failure modes of
modern technical practice. The fifty problems at section~9.10
are the deliverable that turns the volume's investment into the
installed habit.

\subsection{Closing principle: estimation as the engineer's moral
floor}

Volume I closes with a principle that ties together the chapter's
procedural rules, the volume's measurement discipline, and the
forward arc into Volume XII's closing chapter on engineering
ethics. The principle is a single composite statement that the
chapter's apparatus is the engineer's working defence against
the failure mode of confidence without evidence.

\begin{principle}[Estimation is the engineer's moral floor]
The discipline of estimation is the floor below which engineering
work cannot drop without losing its claim to honesty. The
discipline has four working consequences:

\begin{enumerate}\setlength\itemsep{2pt}
\item \emph{Estimation before calculation is honesty.} A recorded
estimate, dated and decomposed, is a blind comparison against
which the later calculation can be checked. The discipline
preserves the distinction between what the engineer expected and
what the calculation produced, and it permits the disagreement
to be investigated rather than absorbed.

\item \emph{Estimation enables critique.} A number reported
without an estimate is a number that resists examination. A
number reported with an estimate is a number whose decomposition,
factors, and stated range invite the reviewer to test each
factor independently. The discipline opens the calculation to
external scrutiny.

\item \emph{Estimation makes the engineer accountable.} An
estimate that turned out wrong, recorded and dated, is a
training signal for the engineer who recorded it. The pattern of
which factors the engineer's intuition systematically over- or
under-estimates is the calibration record the discipline
produces. The Brier-score apparatus at section~9.7 is the formal
form of this accountability; the working ledger of (estimate,
outcome) pairs is the informal form.

\item \emph{Estimation is irreversible.} Once the discipline has
been practised across fifty Fermi problems, the engineer cannot
read a quantitative claim without forming an internal estimate
against which to test it. The internal ruler is built; the
order-of-magnitude scale of any new quantity arrives unbidden
within seconds of the quantity being named; and a published
number that contradicts the internal estimate by more than the
factor-of-three target is flagged automatically. The engineer
who has installed the habit can no longer un-install it.
\end{enumerate}

The forward link is to Volume XII Chapter $14$, where engineering
ethics is examined as a working discipline rather than a
professional formality. Estimation, as practised in this chapter,
is the floor on which the broader ethics rests; an engineer
without the discipline has no working defence against the
failure modes Volume XII names. The Volume I closing project at
section~9.10 is the deliverable that installs the floor.
\end{principle}

\section{The fifty Fermi problems: project and exercises}\pathtag{core}

The exercise set for this chapter is also the volume's closing
project. The reader solves all fifty problems, writes out their
working for each, compares to ground truth where it can be looked
up, and writes a 1500-word reflection on the experience. The
project is the deliverable that turns Volume I from a sequence of
chapters into an installed habit.

\begin{project}[Volume I closing project: a personal estimation portfolio]

\textbf{Track and class.}
Analysis only. Hazard class: none. Tools: paper, pencil,
calculator, the reference data tables under
\repopath{volumes/01-quantity/ch09-discipline-of-estimation/data/},
and internet research access (used only after the reader's
estimate has been recorded).

\textbf{Phase 1 (week 1): twelve problems, paced practice.}
Solve the twelve Physical-scales problems (Fermi 1 to 12). For
each, record the question, the decomposition tree, the per-factor
low/nominal/high inputs, the combination, and (after the estimate is
written down) the ground-truth comparison from a named source. Do
not look up ground truth until the estimate is recorded. Deliverable
at the end of week 1: twelve worked solutions, each on a single
page, with the geometric-mean ratio of estimate to truth at the
bottom.

\textbf{Phase 2 (week 2): nineteen problems, pace acceleration.}
Solve the Urban and Civilization problems (Fermi 13 to 22) and the
Biology and Medicine problems (Fermi 23 to 31). At this stage the
reader records, in addition to the point estimate, a stated
$90\,\%$ confidence interval for each. After ground-truth lookup,
mark each interval as containing truth (1) or not containing truth
(0). The empirical hit-rate of the intervals is the reader's
calibration score for the phase. Deliverable: nineteen solutions,
plus a one-page table of stated confidence and hit-or-miss outcome.

\textbf{Phase 3 (week 3): nineteen problems, the harder set.}
Solve the Technology and Engineering problems (Fermi 32 to 42) and
the Environmental and Economic problems (Fermi 43 to 50). These
include the four deep-uncertainty problems, where the deliverable
is a bounded range rather than a point. Deliverable: nineteen
solutions, with the four deep-uncertainty problems marked and
their bounds explicit.

\textbf{Phase 4 (week 3 end): consolidation and reflection.}
Compute three summary statistics across the forty-six
point-estimate problems: the geometric mean of estimate-to-truth
ratios (the systematic-bias estimator); the geometric standard
deviation of the same ratios (the precision estimator); and the
empirical hit-rate of the $90\,\%$ confidence intervals (the
calibration estimator). Plot the histogram of $\log_{10}(\text{
estimate}/\text{truth})$ across the forty-six problems;
\Cref{fig:vol01:ch09:error-histogram} is the reference shape. Then
write a 1500-word reflection that addresses the five questions in
the deliverables list below.

\textbf{Deliverables.}
\begin{enumerate}[leftmargin=*]
\item Fifty solution pages, one per problem.
\item One summary statistics page (geometric mean, geometric
standard deviation, hit-rate).
\item One histogram of log-ratios.
\item One 1500-word reflection that answers: which problem types
the reader's intuition was best at and why; which it was worst at
and what working knowledge would tighten those estimates; what
habits the reader acquired across Volume I; which anchor points in
the internal ruler need the most repair; what one habit the reader
will carry forward into Volume II as the most useful single move.
\end{enumerate}

\textbf{Success criteria.}
\begin{enumerate}[leftmargin=*]
\item All fifty problems attempted with explicit decomposition,
factor estimates, and combination written out.
\item For the forty-six point-estimate problems, geometric mean of
estimate-to-truth ratios within a factor of two of unity (no large
systematic bias) and geometric standard deviation under a factor
of five (working-engineer precision across domains).
\item For the nineteen Phase-2 calibration problems, empirical
hit-rate of the $90\,\%$ intervals between $0.75$ and $1.0$ (the
range a Brier-score-trained estimator can reach within a month).
\item Four deep-uncertainty problems delivered as bounded ranges,
each with the dominant-uncertainty factor named.
\item Reflection of $1500 \pm 200$ words that identifies at least
two domain-specific anchor points the reader will revise after
the exercise.
\end{enumerate}

\textbf{Time budget.}
About one hour per problem in week 1 (twelve problems, twelve
hours); about forty-five minutes per problem in week 2 (nineteen
problems, fourteen hours); about thirty minutes per problem in
week 3 (nineteen problems, ten hours), plus four hours for the
consolidation and reflection. Total: about forty hours, distributed
across three weeks. The rate accelerates with practice because the
internal ruler tightens.

\textbf{Editorial provision.}
Before release, the editor publishes reference solutions to all
fifty problems, with working, named ground-truth sources, access
dates, units, and ratios to the reference values. The reader
compares only \emph{after} completing their own estimates. A Fermi
problem has defensible and indefensible answers; the reflection's
job is to make the reader's defence explicit. The reference table
under \repopath{volumes/01-quantity/ch09-discipline-of-estimation/data/fermi50\_ground\_truth.csv}
is the editor's working draft of the reference values.

\end{project}

\input{volumes/01-quantity/ch09-discipline-of-estimation/figures/fig-error-histogram.tex}

The fifty problems are grouped by domain. Within each group the
problems are roughly in increasing difficulty.

\subsection*{Physical scales: mass, length, time, energy}\pathtag{core}

\begin{exercise}[\textit{Fermi 1, mass scales}]
Estimate the mass of all human beings alive today, in tonnes,
current as of 2024. State your assumptions for world population,
adult fraction, child fraction, mean adult mass, and mean child
mass. Compare your result to a reference calculation built from a
named population source and a named body-mass source; do not
require a published product of population times mean adult mass.
\end{exercise}

\paragraph{Solution sketch.}
World population $\approx 8.0 \times 10^{9}$ (U.N.\ World
Population Prospects 2024); adult fraction $\approx 0.7$ (children
under $18$ are about $0.3$); mean adult mass $\approx 62
\,\si{\kilo\gram}$ (mean of male and female means across regions,
Walpole 2012); mean child mass $\approx 25 \,\si{\kilo\gram}$. The
sum is $8.0 \times 10^{9} \times (0.7 \times 62 + 0.3 \times 25)
\approx 8.0 \times 10^{9} \times 51 \approx 4.1 \times 10^{11}
\,\si{\kilo\gram} = 4.1 \times 10^{8} \,\si{\tonne}$. Within a
factor of two of the $4 \times 10^{8}$ tonnes reference value
constructed from primary sources.

\begin{exercise}[\textit{Fermi 2, mass scales}]
Estimate the total mass of fresh water held in the Earth's
freshwater lakes, in tonnes. Decompose into the surface area of the
fresh-water lakes, the mean lake depth, and the density of fresh
water. Compare to the published figure for global lake water volume
in any standard hydrology reference.
\end{exercise}

\paragraph{Solution sketch.}
Global lake surface area $\approx 2.7 \times 10^{6} \,\si{\kilo\meter
\squared} \approx 2.7 \times 10^{12} \,\si{\meter\squared}$ (Verpoorter
\emph{et al.}\ 2014); mean lake depth, dominated by deep tectonic
lakes such as Baikal, $\approx 40 \,\si{\meter}$; density of fresh
water $1000 \,\si{\kilo\gram\per\meter\cubed}$. The product is
$2.7 \times 10^{12} \times 40 \times 1000 \approx 1.1 \times 10^{17}
\,\si{\kilo\gram} \approx 10^{14} \,\si{\tonne}$. The Verpoorter
reference value is $1.8 \times 10^{14} \,\si{\tonne}$; within a
factor of two. Dominant uncertainty: mean depth, which is sensitive
to a handful of large lakes.

\begin{exercise}[\textit{Fermi 3, length scales}]
Estimate the total length of the public road network on Earth, in
kilometres, current as of 2020. Decompose into the population, the
road length per person, and any structural correction for
urban-versus-rural distribution. Compare to the figure published by
any international transport database (CIA World Factbook, World Bank,
or equivalent).
\end{exercise}

\paragraph{Solution sketch.}
World population $\approx 7.8 \times 10^{9}$ in 2020; road length
per person varies from about $20 \,\si{\meter}$ (low-income
countries) to about $40 \,\si{\meter}$ (high-income with extensive
rural network); income-weighted world mean $\approx 8 \,\si{\meter}$
per person on the conservative side (the OECD-style countries have a
small share of population but a large share of road). The product is
$7.8 \times 10^{9} \times 8 \approx 6 \times 10^{10} \,\si{\meter}
= 6 \times 10^{7} \,\si{\kilo\meter}$. The CIA World Factbook
totals roughly $6.4 \times 10^{7} \,\si{\kilo\meter}$; within
$10\,\%$. Dominant uncertainty: the meaning of ``public road'' (paved
versus unpaved) varies across reporting authorities.

\begin{exercise}[\textit{Fermi 4, length scales}]
Estimate the total length of capillary blood vessels in a single
adult human, in kilometres. Decompose into the volume of blood, the
fraction of the cardiovascular system in capillaries, the typical
capillary diameter, and the geometry. Compare to the figure cited
in any standard human-physiology textbook.
\end{exercise}

\paragraph{Solution sketch.}
Total blood volume $\approx 5 \,\si{\liter} = 5 \times 10^{-3}
\,\si{\meter\cubed}$; capillary fraction of total blood volume
$\approx 5\,\%$ (Pittman 2011); capillary cross-section radius
$\approx 4 \times 10^{-6} \,\si{\meter}$, so cross-section area
$\approx \pi (4 \times 10^{-6})^{2} \approx 5 \times 10^{-11}
\,\si{\meter\squared}$. Capillary volume $\approx 2.5 \times 10^{-4}
\,\si{\meter\cubed}$; total capillary length $\approx 2.5 \times
10^{-4} / (5 \times 10^{-11}) \approx 5 \times 10^{6} \,\si{\meter}
= 5 \times 10^{3} \,\si{\kilo\meter}$. Textbook figure (Pittman)
is $\sim 10^{5} \,\si{\kilo\meter}$; the estimate is low by an
order of magnitude, dominated by the capillary-volume-fraction
factor, which is closer to $10\,\%$ across the microvascular bed.

\begin{exercise}[\textit{Fermi 5, time scales}]
Estimate the number of seconds in a typical human lifetime, current
as of 2020. Decompose into life expectancy in years, days per year,
and seconds per day. Comment on whether the answer changes
materially when life expectancy is taken at the level of a high-
income country versus a low-income country.
\end{exercise}

\paragraph{Solution sketch.}
World life expectancy $\approx 73$ years; days per year $365.25$;
seconds per day $8.64 \times 10^{4}$. Product:
$73 \times 365.25 \times 8.64 \times 10^{4} \approx 2.3 \times
10^{9} \,\si{\second}$. High-income life expectancy
($\approx 82$ years) gives $2.6 \times 10^{9} \,\si{\second}$;
low-income life expectancy ($\approx 65$ years) gives $2.0 \times
10^{9} \,\si{\second}$. The variation is within a factor of $1.3$,
small at the order-of-magnitude scale. A human lifetime is
approximately $10^{9.4}$ seconds at world average.

\begin{exercise}[\textit{Fermi 6, energy scales}]
Estimate the total kinetic energy of a moving long-haul aircraft at
cruise, in joules. Decompose into the aircraft mass and the cruise
velocity. Compare to the chemical energy in the aircraft's fuel
load.
\end{exercise}

\paragraph{Solution sketch.}
Aircraft mass at cruise (B777-300ER, mid-flight) $\approx 2.5
\times 10^{5} \,\si{\kilo\gram}$; cruise velocity $\approx 250
\,\si{\meter\per\second}$ (Mach 0.85 at altitude). Kinetic
energy: $\tfrac{1}{2} \times 2.5 \times 10^{5} \times 250^{2}
\approx 7.8 \times 10^{9} \,\si{\joule}$. Fuel chemical energy:
fuel load $\approx 1 \times 10^{5} \,\si{\kilo\gram}$ at Jet-A
energy density $43 \,\si{\mega\joule\per\kilo\gram}$ gives
$4.3 \times 10^{12} \,\si{\joule}$, three orders of magnitude
larger. The ratio explains why aircraft cannot recover kinetic
energy as a meaningful contribution to range.

\begin{exercise}[\textit{Fermi 7, energy scales}]
Estimate the total annual primary energy consumption of the world,
in joules, current as of 2022. Decompose into world population,
per-capita primary energy consumption, and seconds in a year.
Compare to the figure published by the International Energy Agency
or the Energy Institute Statistical Review of World Energy.
\end{exercise}

\paragraph{Solution sketch.}
World population $\approx 8.0 \times 10^{9}$; per-capita primary
power (world average) $\approx 2500 \,\si{\watt}$; seconds in a
year $3.15 \times 10^{7}$. Product: $8.0 \times 10^{9} \times 2500
\times 3.15 \times 10^{7} \approx 6.3 \times 10^{20} \,\si{\joule}$.
The Energy Institute Statistical Review for 2022 reports $6.04
\times 10^{20} \,\si{\joule}$ (604 EJ)
\cite{data:ch09-bp-statistical-2024}; within $5\,\%$. The estimate
worked through the per-capita-power factor, which is dominated by
high-income-country use; the result is tight because the seconds-
per-year factor is exact and world population is known to within
a percent.

\begin{exercise}[\textit{Fermi 8, planetary scales}]
Estimate the mass of the Earth's atmosphere, in kilograms. Decompose
into the surface area of the Earth, the column mass per unit area
(ambient pressure divided by gravitational acceleration), and the
working approximation that the column mass is roughly the same
across the surface. Compare to the published figure.
\end{exercise}

\paragraph{Solution sketch.}
Earth surface area $\approx 4\pi R_{\oplus}^{2} = 4\pi (6.37 \times
10^{6})^{2} \approx 5.1 \times 10^{14} \,\si{\meter\squared}$;
column mass per area $= p/g = 1.013 \times 10^{5} / 9.81 \approx
1.03 \times 10^{4} \,\si{\kilo\gram\per\meter\squared}$. Product:
$5.1 \times 10^{14} \times 1.03 \times 10^{4} \approx 5.3 \times
10^{18} \,\si{\kilo\gram}$. Reference value (Trenberth 2005):
$5.15 \times 10^{18} \,\si{\kilo\gram}$. Within $3\,\%$. The
estimate is one of the cleanest in the set because all factors are
known to high precision; this is a calibration anchor for the
reader's internal ruler.

\begin{exercise}[\textit{Fermi 9, planetary scales}]
Estimate the mass of the Earth's oceans, in kilograms. Decompose
into the surface area of the oceans, the mean ocean depth, and the
density of seawater. Compare to the published figure.
\end{exercise}

\paragraph{Solution sketch.}
Ocean surface area $\approx 0.71 \times 5.1 \times 10^{14}
\approx 3.6 \times 10^{14} \,\si{\meter\squared}$; mean ocean
depth $\approx 3700 \,\si{\meter}$; seawater density $\approx
1025 \,\si{\kilo\gram\per\meter\cubed}$. Product: $3.6 \times
10^{14} \times 3700 \times 1025 \approx 1.37 \times 10^{21}
\,\si{\kilo\gram}$. Reference value: $1.4 \times 10^{21}
\,\si{\kilo\gram}$ (IPCC AR6). Within $5\,\%$. Like Fermi 8, the
calculation is dominated by well-known factors.

\begin{exercise}[\textit{Fermi 10, cosmic scales}]
Estimate the number of stars in the Milky Way galaxy. Decompose
into the galaxy's volume, the volume per star in the local stellar
neighbourhood (computed from the typical interstellar distance),
and a structural correction for the difference between the local
neighbourhood and the galactic mean. Compare to the published
figure (typically given as $1$ to $4 \times 10^{11}$).
\end{exercise}

\paragraph{Solution sketch.}
Galaxy disk radius $\approx 1.5 \times 10^{4}$ pc; thickness
$\approx 300$ pc; disk volume $\approx \pi (1.5 \times 10^{4})^{2}
\times 300 \approx 2.1 \times 10^{11}$ pc$^{3}$. Local typical
interstellar distance $\approx 1$ pc, so local volume per star
$\approx 1$ pc$^{3}$. Local-to-galactic structural correction:
the local neighbourhood is at moderate density; the galactic bulge
is denser and the outer disk sparser; structural factor $\approx
0.5$. Estimate: $2.1 \times 10^{11} \times 1 \times 0.5 \approx
10^{11}$ stars. The published range $1$ to $4 \times 10^{11}$
contains the estimate; the order-of-magnitude target is met.

\begin{exercise}[\textit{Fermi 11, cosmic scales}]
Estimate the age of the universe in seconds, current as of 2024.
Decompose into the published age in years and the seconds in a
year. State the bound on uncertainty from the published age's own
uncertainty.
\end{exercise}

\paragraph{Solution sketch.}
Published age (Planck 2018) $\approx 1.378 \times 10^{10}$ years
with a $\pm 0.020 \times 10^{10}$-year one-sigma uncertainty;
seconds per year $3.156 \times 10^{7}$. Product: $1.378 \times
10^{10} \times 3.156 \times 10^{7} \approx 4.35 \times 10^{17}
\,\si{\second}$. The relative uncertainty inherits from the age's
$\pm 1.5\,\%$; the resulting uncertainty band is $\pm 7 \times
10^{15} \,\si{\second}$.

\begin{exercise}[\textit{Fermi 12, atomic scales}]
Estimate the number of atoms in a copper coin of mass $5 \,\si{\gram}$.
Decompose into the coin mass, the atomic mass of copper, and
Avogadro's number. Comment on the precision your estimate
inherits from each factor.
\end{exercise}

\paragraph{Solution sketch.}
Moles of copper: $m / M = 5 \,\si{\gram} / 63.5 \,\si{\gram\per
\mol} \approx 7.87 \times 10^{-2}$ mol. Atoms: $7.87 \times 10^{-2}
\times 6.022 \times 10^{23} \approx 4.74 \times 10^{22}$ atoms. The
coin mass is the loosest factor (a US cent is now zinc-clad copper
of about $2.5 \,\si{\gram}$; a pure $5 \,\si{\gram}$ copper coin
might be an idealised reference). Atomic mass and Avogadro's number
are exact in the post-2019 SI; the estimate inherits its precision
entirely from the mass and the purity assumption.

\subsection*{Urban and civilization scales}\pathtag{core}

\begin{exercise}[\textit{Fermi 13, urban populations}]
Estimate the number of people who live within $20 \,\si{\kilo\meter}$
of the centre of the largest city in your country. Decompose by area,
mean population density, and any geographic exclusions. Compare to
published metropolitan-area population figures.
\end{exercise}

\paragraph{Solution sketch.}
A circle of radius $20 \,\si{\kilo\meter}$ has area $\pi \times
20^{2} \approx 1257 \,\si{\kilo\meter\squared}$. For a typical
large city in a high-income country (mean metro density of about
$2000 \,\si{persons\per\kilo\meter\squared}$), the count is
$\approx 1257 \times 2000 \approx 2.5 \times 10^{6}$. For a
megacity such as Tokyo or Mumbai (density $\approx 10^{4}$
persons/$\,\si{\kilo\meter\squared}$ within $20 \,\si{\kilo\meter}$),
the count rises to $\approx 1.2 \times 10^{7}$. Reader compares to
the named metropolitan figure; ratio typically within a factor
of two of the central-area census.

\begin{exercise}[\textit{Fermi 14, urban infrastructure}]
Estimate the total length of underground sewer pipe in a metropolitan
area of about $5$ million people. Decompose into the population, the
sewer-pipe length per person, and any structural correction for
density of the urban area. Compare where local utility statistics are
available.
\end{exercise}

\paragraph{Solution sketch.}
Population: $5 \times 10^{6}$. Sewer length per person in a typical
North American metro is $\approx 3 \,\si{\meter}$ per person
(local-collector plus laterals plus trunk); high-density European
cities $\approx 1.5 \,\si{\meter}$ per person; low-density US Sun
Belt $\approx 5 \,\si{\meter}$ per person. Working mid-value: $3
\,\si{\meter}$ per person. Total length: $5 \times 10^{6} \times 3
\approx 1.5 \times 10^{7} \,\si{\meter} = 15{,}000 \,\si{\kilo\meter}$.
A reference comparable is the London sewer network: about $9000
\,\si{\kilo\meter}$ for $9 \times 10^{6}$ residents (i.e.\ $1$ m per
person). The estimate lies within a factor of two of these
references; the density-correction factor is the dominant lever.

\begin{exercise}[\textit{Fermi 15, buildings}]
Estimate the total floor area of all the residential buildings in a
metropolitan area of about $5$ million people, in square metres.
Decompose into the population, the mean residential floor area per
person, and the structural correction for vacancy. Compare to
published municipal land-use statistics.
\end{exercise}

\paragraph{Solution sketch.}
Population: $5 \times 10^{6}$; mean residential floor area per
person $\approx 40 \,\si{\meter\squared}$ in a developed metro (US:
$\approx 70$; EU: $\approx 35$; Asian high-density: $\approx 20$);
vacancy correction $\approx 1.1$. Floor area: $5 \times 10^{6}
\times 40 \times 1.1 \approx 2.2 \times 10^{8} \,\si{\meter\squared}$.
Comparison: New York City reports approximately $2.6 \times 10^{8}
\,\si{\meter\squared}$ of residential floor area for $\approx 8.4
\times 10^{6}$ residents (Department of City Planning PLUTO data).
Order-of-magnitude consistent; the per-person factor is the
dominant lever and varies by a factor of three across countries.

\begin{exercise}[\textit{Fermi 16, traffic}]
Estimate the number of motor vehicles registered in a country of
about $50$ million people. Decompose into the population, the mean
vehicles per person, and the structural correction for urban-rural
distribution. Compare to the country's transport ministry's
published figures.
\end{exercise}

\paragraph{Solution sketch.}
Population: $5 \times 10^{7}$. Vehicles per person in a
high-income country $\approx 0.6$ (cars plus light trucks, OECD
average); in a middle-income country $\approx 0.2$. Working value
$\approx 0.5$. Vehicles registered: $5 \times 10^{7} \times 0.5
\approx 2.5 \times 10^{7}$. For South Korea ($\approx 5.1 \times
10^{7}$ population), the published registration is $\approx 2.5
\times 10^{7}$ as of 2022; within $5\,\%$. For Italy ($\approx 5.9
\times 10^{7}$ population), the figure is $\approx 4.0 \times
10^{7}$; the estimate is low by a factor of $1.6$ because Italian
vehicle ownership per capita is higher than the OECD mean.

\begin{exercise}[\textit{Fermi 17, food}]
Estimate the total mass of food consumed in your country in one
year, in tonnes. Decompose into the population, the daily food
intake per person (in kilocalories or in mass), and the days in a
year. Compare to the figure published by the country's agriculture
ministry or by the FAO.
\end{exercise}

\paragraph{Solution sketch.}
For a country of $5 \times 10^{7}$ people, daily food intake $\approx
1.5 \,\si{\kilo\gram}$ per person per day (water-bearing solids
plus beverages, retail-equivalent), times $365$ days. Total: $5
\times 10^{7} \times 1.5 \times 365 \approx 2.7 \times 10^{10}
\,\si{\kilo\gram} = 2.7 \times 10^{7} \,\si{\tonne}$ per year. FAO
food-balance-sheet totals for Spain ($\approx 4.8 \times 10^{7}$
people, 2020) report approximately $3.2 \times 10^{7}$ tonnes; the
estimate matches within $20\,\%$. Dominant uncertainty: the
distinction between food-purchased and food-eaten (waste correction
$\approx 1.3$).

\begin{exercise}[\textit{Fermi 18, water}]
Estimate the daily domestic water consumption in your country, in
cubic metres. Decompose into the population, the per-capita
domestic water use, and any structural correction. Compare to the
figure published by the country's water utility.
\end{exercise}

\paragraph{Solution sketch.}
For a country of $5 \times 10^{7}$ people, per-capita domestic
water use $\approx 150 \,\si{\liter}$ per person per day (Europe;
USA $\approx 300$; developing $\approx 50$); total daily volume
$5 \times 10^{7} \times 0.15 \approx 7.5 \times 10^{6}
\,\si{\meter\cubed}$ per day. UK national water statistics (Ofwat,
$\approx 6.6 \times 10^{7}$ population) report $\approx 8.5 \times
10^{6} \,\si{\meter\cubed}$ per day domestic; consistent within
$15\,\%$ after scaling for population. Dominant uncertainty:
per-capita use, which varies by a factor of six across countries.

\begin{exercise}[\textit{Fermi 19, energy}]
Estimate the total electrical energy consumed in your country in
one year, in terawatt hours. Decompose into the population, the
per-capita electricity consumption, and the seconds-or-hours in a
year. Compare to the figure published by the country's grid
operator or by the IEA.
\end{exercise}

\paragraph{Solution sketch.}
For a high-income country of $5 \times 10^{7}$, per-capita electric
power $\approx 800 \,\si{\watt}$ (annual mean, all sectors); hours
in a year $8760$; product: $5 \times 10^{7} \times 800 \times 8760
/ 10^{9} \approx 350 \,\text{TWh}$ per year. South Korea (population
$5.1 \times 10^{7}$, 2022) consumed approximately $540 \,\text{TWh}$;
per-capita power $\approx 1200 \,\si{\watt}$ because of large
industrial demand. The estimate is low by a factor of $1.5$;
industrial-intensity correction is the dominant lever.

\begin{exercise}[\textit{Fermi 20, urban transport}]
Estimate the total number of train journeys made in a metropolitan
rail system of about $5$ million people in one year. Decompose into
the population, the journeys per person per year, and the
structural correction for the share of journeys made by rail
versus road. Compare to the figure published by the rail operator.
\end{exercise}

\paragraph{Solution sketch.}
Population: $5 \times 10^{6}$; rail trips per person per year
$\approx 200$ in a transit-oriented metro (London, Tokyo)
versus $\approx 20$ in a sprawl metro (Houston). Working mid-value:
$120$. Total: $5 \times 10^{6} \times 120 = 6 \times 10^{8}$
journeys per year. Reference comparable: Madrid Metro carried
about $7 \times 10^{8}$ journeys in 2019 for a service area of
$\approx 6$ million people; within $15\,\%$. Dominant uncertainty:
the per-person-per-year factor, which depends on the rail system's
modal share.

\begin{exercise}[\textit{Fermi 21, civic services}]
Estimate the number of public-school teachers in a country of $50$
million people. Decompose into the population, the fraction of
school-age children, the mean class size, and the working hours
ratio. Compare to the figure published by the country's education
ministry.
\end{exercise}

\paragraph{Solution sketch.}
Population: $5 \times 10^{7}$; school-age fraction (ages 5--17)
$\approx 0.18$; thus $\approx 9 \times 10^{6}$ school-age
children. Mean class size $\approx 22$; teachers $\approx 9
\times 10^{6} / 22 \approx 4.1 \times 10^{5}$. Working-hours
correction (one teacher serves multiple classes per day, but
the metric is one teacher per class-load equivalent): factor
$\approx 1.1$ for support staff. Final estimate: $\approx 4.5
\times 10^{5}$ teachers. UK schools workforce census (population
$\approx 5.6 \times 10^{7}$): approximately $5 \times 10^{5}$
teachers; within $15\,\%$ after population scaling.

\begin{exercise}[\textit{Fermi 22, civic infrastructure}]
Estimate the total mass of concrete poured in your country in the
last calendar year, in tonnes. Decompose into population, per-capita
annual concrete consumption, and a structural correction for
construction intensity. If the published source reports cement
rather than concrete, state the assumed cement fraction in concrete
and convert explicitly. Compare to the figure published by the
national cement or ready-mix concrete association.
\end{exercise}

\paragraph{Solution sketch.}
For a country of $5 \times 10^{7}$ people, world-mean per-capita
annual cement consumption $\approx 500 \,\si{\kilo\gram}$ per
person per year; concrete is about $13\,\%$ cement by mass, so
per-capita concrete $\approx 4 \,\si{\tonne}$ per year. Total:
$5 \times 10^{7} \times 4 \approx 2.0 \times 10^{8}$ tonnes per
year. For Spain ($\approx 4.7 \times 10^{7}$, mid-construction
cycle), Cembureau reports cement use of $\approx 1.4 \times 10^{7}$
tonnes; concrete equivalent $\approx 1.1 \times 10^{8}$ tonnes.
The estimate is high by a factor of two because per-capita cement
in mature economies is below the world mean; China and India
dominate the world figure.

\subsection*{Biology and medicine}\pathtag{core}

\begin{exercise}[\textit{Fermi 23, body composition}]
Estimate the total mass of water in a typical adult human, in
kilograms. Decompose into the adult mass and the typical water
fraction (roughly $0.6$). Compare to the published figure from any
human-physiology reference.
\end{exercise}

\paragraph{Solution sketch.}
Adult mass $70 \,\si{\kilo\gram}$; water fraction $0.6$ (a young
adult; the fraction drops with age and adipose mass). Water mass:
$42 \,\si{\kilo\gram}$. Reference value (any standard physiology
text, e.g.\ Tortora): $60 \pm 10 \,\%$ body water for typical
adults. The estimate sits in the middle of the published band; the
factor is well anchored.

\begin{exercise}[\textit{Fermi 24, body composition}]
Estimate the total number of cells in a typical adult human.
Decompose into the typical cell mass (a useful anchor point is
$10^{-9}$ to $10^{-12}$ kilograms depending on cell type) and the
adult mass. Compare to the figure published in any standard cell-
biology textbook (the typical answer cited is $30$ to $40$
trillion cells).
\end{exercise}

\paragraph{Solution sketch.}
A naive estimate using a typical eukaryotic cell mass of
$10^{-12} \,\si{\kilo\gram}$ for a $70 \,\si{\kilo\gram}$ adult
gives $7 \times 10^{13}$ cells, which is too high. A more careful
estimate weights by cell type: red blood cells dominate the count
($\approx 2.5 \times 10^{13}$ in $5 \,\si{\liter}$ of blood at
$5 \times 10^{9}$ per mL); other cells contribute roughly the same
order. Sender, Fuchs, and Milo (2016) carried out the careful
weighting and arrived at $3.0 \times 10^{13}$ human cells, close
to the textbook ``$30$--$40$ trillion'' band
\cite{paper:ch09-sender-cells-2016}. The naive mass-divided-by-cell-mass
estimate is high by about a factor of $2$; the lesson is that
``typical cell mass'' hides a wide distribution.

\begin{exercise}[\textit{Fermi 25, metabolism}]
Estimate the total energy a typical adult human expends in one day,
in kilojoules. Decompose into the basal metabolic rate per
kilogram, the adult mass, the activity factor (a working anchor
point is $1.5$ for moderate activity), and the seconds per day.
Compare to the figure published by any human-nutrition reference.
\end{exercise}

\paragraph{Solution sketch.}
Basal metabolic rate $\approx 1 \,\si{\watt\per\kilo\gram}$; for a
$70 \,\si{\kilo\gram}$ adult, BMR $\approx 70 \,\si{\watt}$.
Activity factor $1.5$ gives $105 \,\si{\watt}$. Day length $8.64
\times 10^{4} \,\si{\second}$. Daily energy: $105 \times 8.64
\times 10^{4} \approx 9.1 \times 10^{6} \,\si{\joule} \approx
9100 \,\si{\kilo\joule}$, equivalent to $\approx 2200 \,\text{kcal}$.
The published recommended daily intake of $2000$--$2500 \,\text{kcal}$
for moderate activity contains the estimate.

\begin{exercise}[\textit{Fermi 26, breathing}]
Estimate the volume of air a typical adult human breathes in one
year, in cubic metres. Decompose into the tidal volume, the
respiratory rate, and the seconds in a year. Compare to the figure
published by any human-physiology reference.
\end{exercise}

\paragraph{Solution sketch.}
Tidal volume $\approx 0.5 \,\si{\liter} = 5 \times 10^{-4}
\,\si{\meter\cubed}$; respiratory rate $\approx 14$ breaths per
minute $= 0.23$ per second; seconds per year $3.15 \times 10^{7}$.
Annual volume: $5 \times 10^{-4} \times 0.23 \times 3.15 \times
10^{7} \approx 3600 \,\si{\meter\cubed}$. The published EPA
standard for residential air exposure is $\approx 4500$
$\si{\meter\cubed}$ per year (Exposure Factors Handbook, 2011);
within $25\,\%$. Difference: the EPA figure includes a higher
mean ventilation rate that accounts for activity.

\begin{exercise}[\textit{Fermi 27, blood}]
Estimate the total volume of blood pumped by the heart in a single
human lifetime, in cubic metres. Decompose into the cardiac output
per minute, the seconds in a year, and the lifespan in years.
Compare to the figure cited in any standard physiology textbook.
\end{exercise}

\paragraph{Solution sketch.}
Cardiac output at rest $\approx 5 \,\si{\liter\per\min} = 8.3
\times 10^{-5} \,\si{\meter\cubed\per\second}$; lifespan $80$
years $= 2.5 \times 10^{9} \,\si{\second}$. Total volume:
$8.3 \times 10^{-5} \times 2.5 \times 10^{9} \approx 2.1 \times
10^{5} \,\si{\meter\cubed}$. The textbook figure is approximately
$2 \times 10^{5} \,\si{\meter\cubed}$ (about $200{,}000$ cubic
metres); within $5\,\%$. Anchor point: $5 \,\si{\liter}$ of blood
pass through the heart at rest every minute for a lifetime.

\begin{exercise}[\textit{Fermi 28, neurons}]
Estimate the number of synapses in a typical adult human brain.
Decompose into the number of neurons (a working anchor point is
$10^{11}$) and the synapses per neuron (a working anchor point is
$10^{3}$ to $10^{4}$). Compare to the figure cited in any standard
neuroscience reference.
\end{exercise}

\paragraph{Solution sketch.}
Neurons $\approx 10^{11}$ (Herculano-Houzel 2009 revised from
Kandel's earlier $10^{12}$); mean synapses per neuron $\approx 5
\times 10^{3}$. Total: $10^{11} \times 5 \times 10^{3} = 5 \times
10^{14}$ synapses. The standard neuroscience reference range is
$1 \times 10^{14}$ to $5 \times 10^{14}$; the estimate sits at
the upper end. Dominant uncertainty: synapses per neuron varies
across cortical areas by a factor of three.

\begin{exercise}[\textit{Fermi 29, microbes}]
Estimate the number of bacterial cells in the human gut. Decompose
into the typical bacterial-cell mass and the typical mass of the
gut microbiome. Compare to the figure published in the bacterial-
to-human-cell ratio literature (roughly $10^{13}$ to $10^{14}$).
\end{exercise}

\paragraph{Solution sketch.}
Bacterial cell mass $\approx 10^{-15} \,\si{\kilo\gram}$ (E.\ coli
reference); gut microbiome mass $\approx 0.2 \,\si{\kilo\gram}$
(estimate widely circulated); naive count $\approx 0.2 / 10^{-15}
= 2 \times 10^{14}$ cells. Sender, Fuchs, and Milo's (2016)
careful weighting gives $3.8 \times 10^{13}$ bacteria in a
$70 \,\si{\kilo\gram}$ adult \cite{paper:ch09-sender-cells-2016};
within a factor of five of the naive estimate. The earlier $10:1$
bacteria-to-human-cell ratio claim was an overestimate by about
$10\times$, corrected by the Sender paper to about $1:1$.

\begin{exercise}[\textit{Fermi 30, drug doses}]
Estimate the total mass of acetaminophen (paracetamol) consumed
worldwide in one year, in tonnes. Decompose into the world
population, the fraction taking the drug regularly, the fraction
taking it occasionally, and the typical annual consumption per user.
Compare where pharmaceutical-industry production statistics are
available.
\end{exercise}

\paragraph{Solution sketch.}
World population: $8.0 \times 10^{9}$; fraction with regular access
to paracetamol $\approx 0.4$ (high- and middle-income countries);
mean doses per person per year $\approx 10$ (occasional use); dose
mass $\approx 500 \,\si{\milli\gram}$. Total: $8 \times 10^{9}
\times 0.4 \times 10 \times 0.5 \,\si{\gram} = 1.6 \times 10^{10}
\,\si{\gram} = 1.6 \times 10^{4} \,\si{\tonne}$. Industry estimates
of world paracetamol production are in the range $1 \times 10^{5}$
to $2 \times 10^{5}$ tonnes per year (active pharmaceutical
ingredient). The estimate is low by about an order of magnitude;
dominant uncertainty is the per-user annual consumption, which is
much higher for chronic-pain users.

\begin{exercise}[\textit{Fermi 31, healthcare}]
Estimate the number of medical X-ray examinations performed in a
country of $50$ million people in one year. Decompose into the
population, the X-ray rate per person per year, and any structural
correction. Compare to the figure published by the country's
radiation-protection authority.
\end{exercise}

\paragraph{Solution sketch.}
Population: $5 \times 10^{7}$; X-ray rate $\approx 1$ per person
per year (UNSCEAR global average for high-income countries); total
$\approx 5 \times 10^{7}$. UK Health Protection Agency reports
approximately $4.5 \times 10^{7}$ X-ray procedures per year on a
population of $\approx 6.6 \times 10^{7}$; per-capita rate
$\approx 0.68$. The estimate is high by a factor of $1.5$, within
the working factor-of-two target. The rate varies by a factor of
three across countries (higher in the USA, lower in middle-income
settings).

\begin{exercise}[\textit{Fermi 32, public health}]
Estimate the total number of doses of seasonal influenza vaccine
administered in a country of $50$ million people in one year.
Decompose into the population, the vaccination coverage rate, and
the structural correction for repeat dosing. Compare to the figure
published by the country's public-health agency.
\end{exercise}

\paragraph{Solution sketch.}
Population: $5 \times 10^{7}$; vaccination coverage rate $\approx
0.3$ (typical for adult plus elderly programmes); repeat dosing
correction $\approx 1$ (one dose per year for most adults). Total:
$5 \times 10^{7} \times 0.3 \approx 1.5 \times 10^{7}$ doses.
England's Joint Committee on Vaccination and Immunisation reports
approximately $1.8 \times 10^{7}$ flu vaccines for the 2019--2020
season on a population of $\approx 5.6 \times 10^{7}$; per-capita
coverage $0.32$. The estimate is within $5\,\%$ after scaling.

\subsection*{Technology and engineering}\pathtag{core}

\begin{exercise}[\textit{Fermi 33, computation}]
Estimate the total number of transistors fabricated worldwide in
one year, current as of 2022. Decompose into wafer starts, die per
wafer, transistors per die, and chip mix. Treat the comparison as
an industry-estimate problem: name the source used for wafer starts
and the source used for representative transistor counts, then
report the answer as an order-of-magnitude range rather than as a
single published truth.
\end{exercise}

\paragraph{Solution sketch.}
Wafer starts (SEMI, 2022): $\approx 1.5 \times 10^{8}$
$300\,\si{\milli\meter}$-equivalent wafers; usable die area per
wafer $\approx 7 \times 10^{4} \,\si{\milli\meter\squared}$;
representative leading-edge die transistor density $\approx 10^{8}$
per $\,\si{\milli\meter\squared}$, but trailing-edge nodes
(automotive, power) sit at $\approx 10^{6}$ per $\,\si{\milli\meter\squared}$.
Weighted mean $\approx 10^{7}$ per $\,\si{\milli\meter\squared}$.
Total: $1.5 \times 10^{8} \times 7 \times 10^{4} \times 10^{7}
\approx 10^{20}$ transistors per year. Range: $10^{20}$ to $10^{22}$
depending on the mix weighting. Published estimates by Koomey and
others cluster around $10^{22}$ transistors per year as of 2022.

\begin{exercise}[\textit{Fermi 34, computation}]
Estimate the total electrical energy consumed by all the world's
data centres in one year, in terawatt hours, current as of 2022.
Decompose into the world's data-centre electrical power demand and
the seconds in a year. Compare to the figure published by the IEA
(typical answer is around $200$ to $400$ TWh per year as of 2022).
\end{exercise}

\paragraph{Solution sketch.}
World data-centre electrical power demand: world internet-connected
device count $\approx 5 \times 10^{9}$, with $\approx 30 \,\si{\watt}$
of data-centre power allocated per device on average (hyperscale
plus enterprise plus edge). Total power $\approx 1.5 \times 10^{11}
\,\si{\watt}$; hours per year $8760$; annual energy $\approx 1.3
\times 10^{12} \,\text{kWh} = 1300 \,\text{TWh}$. The IEA 2024
range for 2022 is $240$ to $340 \,\text{TWh}$
\cite{data:ch09-iea-data-centres-2024}; the estimate is high by a
factor of about $4$, dominated by an aggressive
power-per-device allocation. A working anchor: data-centre power
demand is approximately $1\,\%$ of world electricity, and world
electricity is $\approx 2.9 \times 10^{4} \,\text{TWh}$, so
$\approx 290 \,\text{TWh}$ is the right order.

\begin{exercise}[\textit{Fermi 35, networking}]\pathtag{standard}
Estimate the total volume of data transmitted across the global
internet in one second, in bytes per second, current as of 2024.
Decompose into annual global internet traffic or user count,
per-user traffic, and duty cycle. Compare to a named current source
such as ITU, TeleGeography, Cloudflare, or a current Cisco archived
forecast only if the year and metric match.
\end{exercise}

\paragraph{Solution sketch.}
User count $\approx 5.4 \times 10^{9}$ (ITU, 2024); per-user
average traffic $\approx 200 \,\si{\giga\byte}$ per month $= 6.4
\,\si{\giga\byte}$ per day; spread evenly over the day this is
$\approx 7.4 \times 10^{4} \,\si{\byte\per\second}$. Total:
$5.4 \times 10^{9} \times 7.4 \times 10^{4} \approx 4 \times 10^{14}
\,\si{\byte\per\second}$. Duty-cycle correction (peak versus
average) increases the peak rate by a factor of $\approx 2$. The
TeleGeography 2024 estimate of global IP traffic averages
approximately $1$ to $2 \times 10^{14} \,\si{\byte\per\second}$;
the estimate is high by a factor of two to three, dominated by
the per-user-traffic figure (skewed by video streaming and CDN
caching).

\begin{exercise}[\textit{Fermi 36, manufacturing}]
Estimate the total number of motor cars manufactured worldwide in
one year, current as of 2022. Decompose into the world population,
the rate of car replacement per person per year, and the structural
correction. Compare to the figure published by the International
Organization of Motor Vehicle Manufacturers (OICA).
\end{exercise}

\paragraph{Solution sketch.}
World population: $8.0 \times 10^{9}$; world car-ownership rate
$\approx 0.2$ cars per person (heavily weighted by high-income
countries); cars in service $\approx 1.6 \times 10^{9}$; mean
service life $\approx 15$ years; replacement rate $\approx
1.6 \times 10^{9} / 15 \approx 1.1 \times 10^{8}$ cars per year.
With a growth correction of $+10\,\%$ for fleet expansion, the
estimate is $\approx 7$ to $8 \times 10^{7}$ passenger cars per
year. OICA reports $6.7 \times 10^{7}$ passenger cars produced in
2022 \cite{data:ch09-oica-2023}; within $20\,\%$.

\begin{exercise}[\textit{Fermi 37, aerospace}]
Estimate the total number of commercial-aviation passenger seats in
service worldwide, current as of 2022. Decompose into the size of
the global commercial fleet, the typical seats per aircraft, and
the structural correction for aircraft mix (narrowbody vs.~widebody
vs.~regional). Compare to the figure published by IATA.
\end{exercise}

\paragraph{Solution sketch.}
Global commercial passenger fleet (post-pandemic, 2022) $\approx
2.5 \times 10^{4}$ aircraft; typical seat count weighted across mix
$\approx 170$ (single-aisle dominant, with $\approx 150$ seats;
twin-aisle $\approx 300$; regional $\approx 70$). Total seats:
$2.5 \times 10^{4} \times 170 \approx 4.3 \times 10^{6}$. IATA
reports approximately $4.0 \times 10^{6}$ seats in service in
2022; within $10\,\%$.

\begin{exercise}[\textit{Fermi 38, software}]\pathtag{standard}
Estimate the number of lines of source code in one named open-source
operating-system component, such as the Linux kernel, current as of
2024. Decompose into subsystem count, files per subsystem, lines
per file, and generated-code exclusions. Compare to a reproducible
source-line count from the release tarball or repository. Do not
use uncitable Windows or macOS line-count folklore as ground truth.
\end{exercise}

\paragraph{Solution sketch.}
Linux kernel top-level directories (subsystems): $\approx 30$
(drivers, fs, net, mm, arch, kernel, etc.). Mean files per
subsystem $\approx 2000$ (drivers dominates with $\approx 60{,}000$);
mean lines per file $\approx 200$. Total: $30 \times 2000 \times
200 \approx 1.2 \times 10^{7}$ lines. The reproducible
\texttt{cloc}-style count on Linux $6.5$ (October $2023$) returns
approximately $2.8 \times 10^{7}$ source lines, dominated by the
drivers subtree at $\approx 1.5 \times 10^{7}$ lines. The estimate
is low by a factor of two; the working-knowledge correction is
that the drivers tree is several times larger than a per-subsystem
mean would suggest.

\begin{exercise}[\textit{Fermi 39, sensors}]
Estimate the total number of accelerometer readings produced by all
smartphones in the world in one second. Decompose into installed
smartphone count, fraction of phones sampling at that instant,
sampling rate, and whether the phone stores, discards, or transmits
the samples. Treat this as a no-ground-truth Fermi problem unless
the reference solution supplies a named industry data source; the
postmortem should identify the sampling-duty-cycle assumption as
the dominant uncertainty.
\end{exercise}

\paragraph{Solution sketch.}
Installed smartphones $\approx 6 \times 10^{9}$; fraction sampling
at any moment (screen on or background sensor active) $\approx
0.1$; nominal sampling rate when active $\approx 100 \,\si{\hertz}$;
samples per second produced: $6 \times 10^{9} \times 0.1 \times
100 = 6 \times 10^{10}$. Most are discarded immediately by sensor
fusion; only a small fraction is stored or transmitted. The
ratio of generated to stored samples is the dominant uncertainty,
of order $10^{3}$. The estimate is a no-ground-truth Fermi
problem; the lesson is the dominance of duty-cycle assumptions.

\begin{exercise}[\textit{Fermi 40, manufacturing}]
Estimate the total mass of plastic produced worldwide in one year,
in tonnes, current as of 2022. Decompose into the world population,
the per-capita annual plastic consumption, and any structural
correction. Compare to the figure published by Plastics Europe or
similar industry sources (typical answer is around $400$ million
tonnes).
\end{exercise}

\paragraph{Solution sketch.}
This is the second canonical estimate (see section 9.5.4): world
population $8.0 \times 10^{9}$; per-capita plastic use $\approx
50 \,\si{\kilo\gram}$ per person per year; total $\approx 4 \times
10^{8}$ tonnes per year. Plastics Europe 2023 reports approximately
$4 \times 10^{8}$ tonnes for 2022 \cite{data:ch09-plastics-europe-2023};
the estimate matches the published value within $10\,\%$.

\begin{exercise}[\textit{Fermi 41, manufacturing}]
Estimate the total mass of steel produced worldwide in one year, in
tonnes, current as of 2022. Decompose into the world population,
the per-capita annual steel consumption, and any structural
correction. Compare to the figure published by the World Steel
Association.
\end{exercise}

\paragraph{Solution sketch.}
World population $8.0 \times 10^{9}$; per-capita annual steel
$\approx 230 \,\si{\kilo\gram}$ per person per year (world average
heavily weighted by China at $\approx 700 \,\si{\kilo\gram}$ per
person and OECD at $\approx 450 \,\si{\kilo\gram}$ per person).
Total: $8.0 \times 10^{9} \times 230 \approx 1.85 \times 10^{9}$
tonnes per year. World Steel Association reports $1.89 \times
10^{9}$ tonnes of crude steel for 2022 \cite{data:ch09-worldsteel-2023};
within $5\,\%$.

\begin{exercise}[\textit{Fermi 42, computation}]\pathtag{standard}
Estimate the total floating-point operations per second supplied by
the world's top $500$ supercomputers, current as of 2024. Decompose
into the typical performance of the top machine, the slope of the
top-500 list, and the integral over the list. Compare to the figure
published in the TOP500 list.
\end{exercise}

\paragraph{Solution sketch.}
Top machine (Frontier, ORNL, 2024) $\approx 1.2 \times 10^{18}$
$\,\text{FLOP/s}$; the TOP500 list falls roughly as $r^{-1}$ from
the top (one decade per $\approx 100$ entries by approximate
power-law fit). Integral approximation: $\int_{1}^{500}
1.2 \times 10^{18} r^{-1} \,dr = 1.2 \times 10^{18} \ln 500
\approx 7.5 \times 10^{18}$ $\,\text{FLOP/s}$. The TOP500 list
November 2024 aggregate is approximately $8.2 \times 10^{18}$
$\,\text{FLOP/s}$; within $10\,\%$.

\subsection*{Environmental and economic scales}\pathtag{core}

\begin{exercise}[\textit{Fermi 43, climate}]
Estimate the total mass of carbon dioxide emitted globally from
fossil-fuel combustion in one year, in tonnes, current as of 2022.
Decompose into the world primary-energy consumption, the fraction
from fossil fuels, the carbon intensity of fossil fuels, and the
seconds in a year. Compare to the figure published by the IEA, the
Global Carbon Project, or the EDGAR database.
\end{exercise}

\paragraph{Solution sketch.}
World primary energy $6 \times 10^{20} \,\si{\joule}$ per year;
fossil-fuel fraction $\approx 0.82$; mean carbon intensity of
fossil fuels $\approx 75 \,\text{g CO}_{2}$ per MJ (mix of coal,
oil, gas). Total: $6 \times 10^{20} \times 0.82 \times 75 \times
10^{-6}\,\si{\gram\per\joule} \approx 3.7 \times 10^{16}
\,\si{\gram} = 3.7 \times 10^{10}$ tonnes of CO$_{2}$ per year.
Global Carbon Project 2023 reports $\approx 3.7 \times 10^{10}$
tonnes for 2022 \cite{data:ch09-gcb-2023}; the estimate matches
within $1\,\%$ thanks to well-calibrated factors.

\begin{exercise}[\textit{Fermi 44, climate}]
Estimate the additional radiative forcing on the Earth produced by
the increase in atmospheric carbon dioxide concentration since
$1850$, in watts per square metre. Decompose into the published
radiative forcing per doubling of carbon dioxide and the
fractional change in concentration. Compare to the figure published
by the IPCC.
\end{exercise}

\paragraph{Solution sketch.}
Radiative forcing per doubling of CO$_{2}$ $\approx 3.7
\,\si{\watt\per\meter\squared}$ (Myhre 1998 formula used by IPCC).
Concentration in 1850 $\approx 285$ ppm; current (2024) $\approx
420$ ppm; ratio $1.47$; log-ratio $\log_{2}(1.47) \approx 0.56$.
Forcing $\approx 3.7 \times 0.56 \approx 2.1
\,\si{\watt\per\meter\squared}$. IPCC AR6 reports $2.16 \pm 0.26
\,\si{\watt\per\meter\squared}$ for CO$_{2}$-only forcing through
2019; within $5\,\%$.

\begin{exercise}[\textit{Fermi 45, environment, deep uncertainty}]\pathtag{standard}
Estimate the total mass of microplastics suspended in the world's
oceans, in tonnes. Acknowledge that this is a deep-uncertainty
problem. Decompose into the annual flux of plastic into the oceans,
the fraction that fragments to microplastic size, the fraction that
remains in suspension, and the integration time. State the upper
and lower bounds defensible from the published literature
explicitly, with the dominant-uncertainty factor identified.
\end{exercise}

\paragraph{Solution sketch.}
Annual plastic flux into oceans $\approx 1 \times 10^{7}$ tonnes
per year (Jambeck 2015 estimate, with uncertainty of a factor of
five); fraction fragmenting to microplastic size over decades
$\approx 0.1$ to $0.5$; suspended fraction $\approx 0.01$ to
$0.1$; integration time $\approx 50$ years (accumulation since
mass-production take-off). Lower bound: $10^{7} \times 0.1 \times
0.01 \times 50 = 5 \times 10^{5}$ tonnes; upper bound: $10^{7}
\times 0.5 \times 0.1 \times 50 = 2.5 \times 10^{7}$ tonnes.
Published literature range spans $1 \times 10^{5}$ to $5 \times
10^{6}$ tonnes for surface microplastics
\cite{paper:zhao2025-ocean-microplastics}. Dominant uncertainty:
the suspended fraction (most plastics sink or wash ashore).

\begin{exercise}[\textit{Fermi 46, environment}]
Estimate the total volume of fresh water withdrawn worldwide for
agricultural irrigation in one year, in cubic metres, current as
of 2020. Decompose into the irrigated area, the mean water
application rate per hectare, and the seconds in a year. Compare
to the figure published by the FAO.
\end{exercise}

\paragraph{Solution sketch.}
World irrigated area $\approx 3.4 \times 10^{6} \,\si{\kilo\meter
\squared} = 3.4 \times 10^{8}$ ha; mean water application rate
$\approx 8000 \,\si{\meter\cubed}$ per ha per year. Total: $3.4
\times 10^{8} \times 8000 \approx 2.7 \times 10^{12}
\,\si{\meter\cubed}$ per year. FAO AQUASTAT for 2020 reports
approximately $2.8 \times 10^{12} \,\si{\meter\cubed}$; within
$5\,\%$.

\begin{exercise}[\textit{Fermi 47, economy}]
Estimate the total annual gross domestic product of the world, in
US dollars, current as of 2022. Decompose into the world population
and the per-capita GDP at purchasing-power parity. Compare to the
figure published by the World Bank or the IMF.
\end{exercise}

\paragraph{Solution sketch.}
World population $8.0 \times 10^{9}$; per-capita GDP PPP $\approx
\$20{,}000$ per person per year (world mean). Total: $8.0 \times
10^{9} \times 20000 \approx \$1.6 \times 10^{14}$. World Bank
reports global GDP PPP for 2022 at $\$1.65 \times 10^{14}$;
within $5\,\%$.

\begin{exercise}[\textit{Fermi 48, economy}]
Estimate the total annual revenue of the global passenger-aviation
industry, in US dollars, current as of 2022. Decompose into the
total passenger-kilometres flown, the average revenue per
passenger-kilometre, and the structural correction for cargo
versus passenger revenue. Compare to the figure published by IATA.
\end{exercise}

\paragraph{Solution sketch.}
Total passenger-kilometres flown (RPK, IATA 2022) $\approx 6 \times
10^{12}$ pax-km; average revenue per passenger-km $\approx \$0.13$;
total passenger revenue $\approx \$8 \times 10^{11}$. Cargo
correction: cargo adds about $20\,\%$ to airline revenue but the
question is passenger only. IATA reports global passenger revenue
$\approx \$5.2 \times 10^{11}$ for 2022 (the pandemic recovery
year); the estimate is high by a factor of $1.5$, dominated by
the unit-revenue figure that fell during pandemic recovery.

\begin{exercise}[\textit{Fermi 49, environment, deep uncertainty}]\pathtag{standard}
Estimate the total mass of soil erosion from agricultural land
worldwide in one year, in tonnes. Acknowledge that this is a
deep-uncertainty problem. Decompose into the global agricultural
area, the mean erosion rate per hectare per year, and the
structural correction for crop and tillage type. State the upper
and lower bounds defensible from the published literature.
\end{exercise}

\paragraph{Solution sketch.}
Global agricultural area $\approx 4.8 \times 10^{9}$ ha (FAO);
mean erosion rate $5$ to $15$ tonnes per ha per year (Borrelli
2017); structural correction for tilled fraction $\approx 0.6$.
Lower bound: $4.8 \times 10^{9} \times 5 \times 0.6 \approx 1.4
\times 10^{10}$ tonnes per year; upper bound: $4.8 \times 10^{9}
\times 15 \times 0.6 \approx 4.3 \times 10^{10}$ tonnes per year.
Borrelli 2017 reports $3.6 \times 10^{10}$ tonnes; within the
range. Dominant uncertainty: the mean erosion rate, which depends
heavily on slope and tillage practice and varies by an order of
magnitude across regions.

\begin{exercise}[\textit{Fermi 50, the closing problem}]\pathtag{mastery}
Estimate the present stock of human-made material, or anthropogenic
mass, in tonnes. Acknowledge that this is a deep-uncertainty problem.
Decompose into major material classes such as concrete, aggregates,
bricks, asphalt, metals, plastics, and glass, then compare to the
anthropogenic-mass literature \cite{paper:elhacham2020-anthropogenic-mass}.
In two or three sentences at the end of your reflection, state what
this number teaches about the material scale of the twelve volumes
ahead.
\end{exercise}

\paragraph{Solution sketch.}
Concrete and aggregates dominate: cumulative cement production
since 1900 $\approx 6 \times 10^{10}$ tonnes, multiplied by the
$7\times$ ratio of concrete-by-mass to cement yields $\approx
4 \times 10^{11}$ tonnes of concrete; aggregates separately
contribute another $\approx 3 \times 10^{11}$ tonnes; bricks
$\approx 9 \times 10^{10}$ tonnes; asphalt $\approx 7 \times
10^{10}$ tonnes; metals (mostly steel) $\approx 4 \times 10^{10}$
tonnes; plastics $\approx 1 \times 10^{10}$ tonnes; glass and
other $\approx 1 \times 10^{10}$ tonnes. Sum: $\approx 9 \times
10^{11}$ to $1.2 \times 10^{12}$ tonnes. Elhacham 2020 reports
that anthropogenic mass surpassed living biomass (1 Tt) around
2020 \cite{paper:elhacham2020-anthropogenic-mass}; the estimate
matches within a factor of $1.5$. The closing reflection:
twelve volumes carry the reader across the physical scale at
which engineered material now equals the living planet's
biomass; estimation discipline is the habit that keeps the
engineer answerable to that scale.
