\chapter{Water: supply, sanitation, climate}
\label{vol12:ch03}

\epigraph{Water is the most political of all infrastructures: every drop has a sender, a receiver, a contested rightful claim, and a downstream consequence.}{Author}

\chapmeta{Half-life: medium-life. Archetypes: transport, scaling, ethics. Exercise target: 25. Project track: simulation (city water resilience assessment). Hazard class: high (public-health stakes; climate exposure). Reader path default: standard, with the supply and sanitation sections core, the climate section core, and the industrial-water and policy sections mastery. Named cases: Cape Town ``day zero'' 2018 (\cite{acc:csir-capetown-2018}); Jakarta land subsidence (\cite{paper:abidin-jakarta-2011}); Flint water crisis 2014-2019 (\cite{acc:fwatf-2016}); Aral Sea (\cite{paper:micklin-aral-2007}).}

\noindent
Water is the second master commodity of civilization, after energy. Every human settlement is located where it is because of water; every city's growth has hit a water ceiling at some point in its history; every climate-vulnerable region is climate-vulnerable substantially because of water. The water-engineering profession is older than any other engineering profession, with roots in Roman aqueducts, Mesopotamian irrigation, and the centuries-long Dutch fight against the sea. The contemporary water-engineering profession operates at planetary scale: $4\,\text{km}^3/\text{yr}$ of global withdrawals, roughly $70\%$ for agriculture, $20\%$ for industry, $10\%$ for municipal use. The chapter takes the water cycle as the unit of analysis and the urban water system as the recurring case.

\section{The water cycle as engineered domain}\pathtag{core}

The hydrological cycle moves roughly $5 \times 10^5\,\text{km}^3/\text{yr}$ of water through the atmosphere via evaporation and precipitation. Of the precipitation that falls on land ($\approx 1.2 \times 10^5\,\text{km}^3/\text{yr}$), roughly $60\%$ returns to the atmosphere through evapotranspiration; the remaining $40\%$ becomes runoff that feeds rivers, lakes, and groundwater. Of that runoff, humanity withdraws roughly $4 \times 10^3\,\text{km}^3/\text{yr}$, or approximately $10\%$ of available freshwater. This withdrawal rate is averaged: in many basins (Indus, Colorado, Yellow River, Murray-Darling), withdrawal exceeds local renewable supply and is being made up by groundwater mining or upstream transfers.

The engineered water system is overlaid on the natural cycle. Reservoirs (some $50{,}000$ large dams worldwide) store roughly $7{,}000\,\text{km}^3$ of water; aqueducts and canals transfer water across thousands of kilometres; treatment plants render natural water potable; distribution networks deliver it; collection networks remove wastewater; treatment plants render it returnable to the environment. The system is mature in the developed world (water reaches the tap, sewage leaves the building, mostly) and developing rapidly elsewhere, but the WHO and UNICEF report (\cite{web:who-unicef-jmp-2023}) that approximately 2.2 billion people lack safely managed drinking water and approximately 3.5 billion lack safely managed sanitation as of 2022.

\begin{archetype}[Scaling: the urban water budget]
A city of 1 million people consumes roughly $200{,}000\,\text{m}^3/\text{day}$ of water at $200\,\text{L/person/day}$ (a typical urban use). The system that delivers this water includes: a source (river, lake, reservoir, aquifer) with multi-year reliability; treatment plants sized for peak day (typically $1.5\times$ the average day); transmission mains; service reservoirs sized for peak hour storage; distribution networks; sewers sized for peak wet-weather flow (typically $5$-$10\times$ the dry-weather flow); wastewater treatment plants; receiving water. The total engineered capacity (water plus wastewater) is approximately $5\times$ the average flow, providing the redundancy for the system to operate through demand peaks and weather events.
\end{archetype}

\section{Drinking-water supply at city scale}\pathtag{core}

Drinking-water supply is the engineering of source, treatment, transmission, storage, and distribution. The source choice is local: most cities use surface water (rivers, lakes, reservoirs); some use groundwater; many use a mix. Surface-water sources require treatment for turbidity, organics, microbial contamination, and dissolved minerals. Groundwater requires treatment for hardness, iron and manganese, sometimes arsenic and other dissolved species. The treatment train is well-established (coagulation, flocculation, sedimentation, filtration, disinfection), with disinfection by chlorination remaining the dominant approach. UV and ozone disinfection are alternatives that avoid disinfection byproducts but lack chlorine's residual protection through the distribution network.

The distribution network is typically the largest single capital asset of a water utility. A city of 1 million has roughly $3{,}000$-$5{,}000\,\text{km}$ of buried pipes; replacement cost approximates $\$10$-$\$20$ billion; average replacement rate at the historical pace of $0.5\%/\text{yr}$ implies a 200-year replacement cycle, against typical pipe lives of 50-100 years. The result: most developed-world cities are carrying substantial pipe-replacement deficits. Non-revenue water (the difference between what is treated and what is billed) averages $20\%$-$30\%$ in many cities and reaches $40\%$-$60\%$ in stressed systems.

\begin{principle}[Distribution-network asset management]
Buried pipes are the most under-managed asset in most water utilities because the replacement cost is enormous, the failure consequences are slow (leakage, occasional bursts), and the political incentive is for new build rather than replacement. The engineering response is asset-management programmes that prioritise replacement by failure consequence (high-pressure, large-diameter, downtown locations first), by failure probability (oldest cast-iron pipes; soils with corrosion history; previous-failure clusters), and by sustained replacement-rate funding rather than reactive repair. Cities that have invested in such programmes (Tokyo, Singapore, parts of Western Europe) achieve sub-$10\%$ non-revenue water. Cities that have not (much of the USA, much of the developing world) carry continuous deficits.
\end{principle}

\begin{estimation}
\textbf{Question.} Estimate the buried-pipe replacement deficit of a US city of 1 million people whose pipe network has an average age of 50 years against a design life of 75 years.

\textbf{Reasoning.} Pipe inventory: $4{,}000\,\text{km}$. Replacement cost: $\$3{,}000$/m typical for urban water mains, so $\approx \$12$ billion for full replacement. With average age 50 years against 75-year design life, $\approx 30\%$ of pipes are past their lifetime midpoint and accumulating failure probability. The proportional under-investment, if the city has replaced $0.5\%/\text{yr}$ against the $1.3\%/\text{yr}$ that the design life requires, accumulates to $\approx \$8$ billion of deferred replacement over 50 years. Annual gap at the current under-investment rate: $\$100$ million per year that the city is not funding. Conclusion: most US water utilities of this size are carrying multi-billion-dollar deferred-replacement liabilities; the political-economic context that produces this outcome is the failure mode that Chapter 14's maintenance-failure analysis described.
\end{estimation}

\section{Sanitation and wastewater}\pathtag{core}

Sanitation is the parallel system to drinking-water supply: collection of wastewater from buildings, transport via gravity sewers (typically) to a treatment plant, treatment to a standard that allows discharge to a receiving water, and management of the residual sludge. The collection network in a city of 1 million has approximately the same scale as the distribution network for water (3,000-5,000 km), with similar deferred-replacement dynamics. Combined sewers (which carry both sanitary sewage and stormwater) are a 19th-century inheritance in many older cities (New York, Chicago, Paris, much of the UK) and produce combined-sewer overflows during heavy rain, releasing untreated sewage to receiving waters; their separation is a multi-decade civil engineering programme.

Wastewater treatment is the engineering of biological processes (activated sludge, trickling filters, sequencing batch reactors, membrane bioreactors) combined with primary clarification and tertiary polishing. The treatment level required depends on the receiving water and regulatory regime: primary treatment removes solids; secondary treatment removes organic matter and ammonia; tertiary treatment removes nitrogen, phosphorus, and pathogens. Modern plants achieve $>95\%$ removal of conventional pollutants and substantial removal of micropollutants (pharmaceuticals, personal-care products, microplastics), with the latter remaining an emerging challenge. Sludge management (the solid residual) is approximately one-third of the operational complexity and cost of a wastewater plant; options include land application, incineration, and landfilling, each with environmental trade-offs.

\section{Industrial water}\pathtag{mastery}

Industrial water use is roughly $20\%$ of global withdrawals but concentrated in a few sectors: thermoelectric cooling, semiconductor manufacturing, mining, pulp and paper, food and beverage. Thermoelectric cooling is the largest single industrial use in many countries and is increasingly constrained by source-water temperature limits as climate warms. Semiconductor manufacturing uses ultrapure water at $10$-$100$ litres per chip, with chip fabs typically requiring $5$-$50$ MGD; the recent Taiwan drought (\cite{paper:liao-taiwan-drought-2022}) revealed the strategic dependence of the global semiconductor supply chain on Taiwan's water system. Mining is the most disruptive industrial water use, with extraction-watershed conflicts in Chile (lithium and copper), Peru (copper, gold), and increasingly in the developing-world rare-earths sector.

The engineering response: water-reuse programmes (semiconductor fabs now achieve $80\%$+ reuse), efficiency improvements (modern thermoelectric plants use $1/10$ the water of older once-through designs), and watershed-management agreements with downstream and upstream stakeholders. Each is an engineering and institutional negotiation.

\section{Agricultural water}\pathtag{standard}

Agricultural water is $70\%$ of global freshwater use and is the largest single source of water-stress generation in most basins. The system is largely irrigation (channelled surface water or pumped groundwater applied to crops) with smaller contributions from livestock watering and aquaculture. Irrigation efficiency varies from $30\%$ (traditional flood irrigation) to $90\%$ (drip irrigation), but ``efficiency'' has a counterintuitive feature at basin scale: water not consumed by the crop returns to the basin as runoff or recharge and is available downstream. The basin-scale consequence of replacing flood irrigation with drip is sometimes a reduction in downstream water availability, even as the irrigated area increases. This is one of the harder engineering-economics insights in water-resources management (\cite{paper:perry-irrigation-2007}).

The agricultural-water failure mode at planetary scale: groundwater overdraft. The Ogallala Aquifer in the US High Plains, the North China Plain aquifer, the Indus and Ganges aquifers, the Murray-Darling basin in Australia are all being mined at unsustainable rates. The civilization-scale consequence: agricultural production currently supported by groundwater mining will not be sustained when the aquifers are depleted. The transition will require irrigation efficiency improvements, crop changes, and in some regions a reduction in irrigated area.

\section{Water under climate change}\pathtag{core}

Climate change is rapidly redistributing the water cycle. The IPCC AR6 \cite{std:ipcc-ar6-wg1-2021} reports that mid-latitude winter precipitation is increasing, subtropical and Mediterranean-climate precipitation is decreasing, snowpack-dependent regions are seeing earlier and reduced snowmelt, and the intensity of precipitation extremes is increasing roughly proportionally to atmospheric moisture content (the Clausius-Clapeyron relation: $+7\%$ per degree Celsius of warming). The engineering implication: water systems designed for the climate of the late 20th century are increasingly mismatched to the climate of the mid-21st century. Flood-control systems sized for historical $100$-year events are seeing those events at $10$-$25$-year intervals. Reservoirs sized for historical snowmelt timing are receiving inflows on different schedules. Drinking-water sources that historically reliably refilled are now reaching deficit conditions in drought years.

The Cape Town ``day zero'' episode of 2018 \cite{acc:csir-capetown-2018} is the clearest contemporary case: a multi-year drought brought the city to within months of running out of water; the response combined emergency demand reduction (per-capita use cut by $50\%$ in months), emergency supply (new desalination, additional groundwater), and a long-term reassessment of the climate basis for the system's design. The episode illustrated the pattern: climate change is forcing water systems to be rebuilt against assumptions that are no longer correct, and the planning horizon is no longer multi-decade-stable but rather a moving target.

\begin{archetype}[Balance: water under climate]
Climate-resilient water systems balance demand management (efficiency, pricing, behavioural change), supply diversification (new sources, desalination, reuse), and storage (reservoirs, aquifer recharge). The optimal balance is region-specific: cities with reliable seasonal precipitation can rely more on storage; cities with chronic stress need more demand management; cities with access to ocean water and energy can use desalination as last-resort supply. The engineering analysis is the cost-and-reliability trade-off across these options; the institutional analysis is the political and equity consequences of each. Cape Town's response combined all three; the same combination would have prevented the crisis if applied a decade earlier.
\end{archetype}

\section{Failure: water failures}\pathtag{core}

The chapter's reference cases each illustrate a distinct failure mode of water engineering at civilization scale.

The Aral Sea collapse \cite{paper:micklin-aral-2007} is the civilization-scale ecological failure: Soviet-era irrigation diverted the Amu Darya and Syr Darya rivers for cotton production; the sea's area decreased from $68{,}000\,\text{km}^2$ in 1960 to less than $10{,}000\,\text{km}^2$ by 2010; the ecosystem and the regional fishery collapsed; the exposed seabed produced a dust-and-salt loading that contaminated the regional agricultural soils and the regional respiratory health. The engineering failure: the irrigation system was designed without accounting for its terminal effect on the receiving water; the institutional failure: the Soviet planning system did not have the channels for the consequences to be surfaced and acted upon.

The Cape Town day-zero episode \cite{acc:csir-capetown-2018} is the climate-mismatched-design failure: a system designed for a climate that no longer obtained came close to delivering zero water to a metropolitan population of $4$ million.

The Flint water crisis \cite{acc:fwatf-2016} is the political-institutional failure: the engineering failure mode (corrosion of lead service lines by under-corrosion-controlled water) was well-understood by the profession but was not protected against by the institutional regime that made the source-water switch. The detailed mechanism and institutional context are documented in the registry entry and discussed at length in Volume X Chapter 14 (maintenance failure) and this chapter.

The Jakarta land subsidence \cite{paper:abidin-jakarta-2011} is the chronic-overuse failure: groundwater extraction has caused parts of the city to subside by more than $4\,\text{m}$ over 40 years, with the subsidence reaching $25\,\text{cm/yr}$ in some neighbourhoods; the city is sinking relative to sea level and is increasingly vulnerable to coastal flooding. The engineering response includes a planned capital relocation (to Nusantara on Borneo), seawall construction, and stricter groundwater regulation; the institutional response is still being worked out.

\begin{principle}[Water as systems-of-systems]
Water is never a single-engineered-system problem. The drinking-water system, the wastewater system, the stormwater system, the groundwater system, the surface-water system, the agricultural-water system, the industrial-water system, and the ecological-water system are all connected. A change to any one propagates through the others. Engineering interventions whose scope ignores this interconnection produce unintended consequences at the same scale as the intended one. The chapter's failure cases all share this pattern; the discipline is to think at the basin scale and the multi-decade horizon rather than at the project scale and the political-cycle horizon.
\end{principle}

\section{Project}\pathtag{core}

\begin{project}[City water resilience assessment]
Choose a city with available water data. Document the water system: sources, treatment, distribution, sanitation, stormwater, wastewater treatment. Document the asset condition: pipe age distribution, plant condition, non-revenue water. Document the climate exposure: projected precipitation change, projected demand change, drought-risk assessment. Assess the resilience to a multi-year drought of the magnitude of Cape Town 2018. Identify the specific failure modes that would occur first; estimate the time-to-failure under continuing current management; identify the interventions that would extend the time-to-failure. Estimate the capital and operational investment required; identify the distributive consequences of the intervention.
\end{project}

\section{Exercises}

\subsection*{Calculation}\pathtag{core}

\begin{exercise}
A city of $2$ million people uses $250\,\text{L/person/day}$. Compute total daily demand in m$^3$, annual demand in m$^3$ and in km$^3$, and compare to a typical urban reservoir capacity of $500$ million m$^3$.
\end{exercise}

\begin{exercise}
A water main of diameter $400\,\text{mm}$, length $1\,\text{km}$, and roughness coefficient $C = 100$ (Hazen-Williams) carries $200\,\text{L/s}$. Compute the head loss across the main using the Hazen-Williams equation. Identify the pumping power required.
\end{exercise}

\begin{exercise}
A wastewater treatment plant treats $50{,}000\,\text{m}^3/\text{day}$ with BOD removal from $300\,\text{mg/L}$ to $20\,\text{mg/L}$. Compute the daily mass of organic matter removed and the daily sludge production at an observed yield of $0.4\,\text{kg SS}/\text{kg BOD}_{\text{removed}}$.
\end{exercise}

\begin{exercise}
A reservoir has live capacity $200\,\text{million m}^3$, mean annual inflow $400\,\text{million m}^3$, and annual demand $300\,\text{million m}^3$. Estimate the probability of refill failure in a year with $30\%$ below-average inflow, assuming initial reservoir level at $50\%$ of live capacity.
\end{exercise}

\begin{exercise}
A drip irrigation system has application efficiency $90\%$ against a flood-irrigation system at $40\%$ application efficiency, both serving a $1{,}000$-ha field with $10{,}000\,\text{m}^3/\text{ha/yr}$ crop water requirement. Compute the on-farm water saving and the basin-scale water saving, assuming $50\%$ of the unconsumed flood-irrigation return flow is recovered downstream.
\end{exercise}

\subsection*{Derivation}\pathtag{standard}

\begin{exercise}
Derive the Clausius-Clapeyron-based scaling of extreme precipitation intensity with global mean temperature. Compare to the IPCC AR6 reported values.
\end{exercise}

\begin{exercise}
Derive the optimal storage-versus-demand-reduction trade-off for a city facing a multi-year drought, given the cost of new reservoir capacity and the cost of demand-reduction programmes.
\end{exercise}

\begin{exercise}
Derive the lead-pipe corrosion rate under different water-chemistry conditions (pH, alkalinity, orthophosphate dosing). Identify the conditions that produced the Flint corrosion event.
\end{exercise}

\begin{exercise}
Derive the relationship between groundwater pumping rate, drawdown, and land subsidence in an unconsolidated aquifer system (Theis equation for drawdown; effective-stress relation for subsidence).
\end{exercise}

\subsection*{Estimation}\pathtag{standard}

\begin{exercise}
Estimate the energy intensity of municipal water supply ($\text{kWh}/\text{m}^3$) for: gravity-fed alpine source; pumped surface water; desalination of seawater; treatment of brackish groundwater. Compare to the energy demand for the total US municipal water sector.
\end{exercise}

\begin{exercise}
Estimate the global capital required to bring safely managed drinking water to the $2.2$ billion people currently without it, at typical capital intensities for developing-world systems.
\end{exercise}

\begin{exercise}
Estimate the embedded water in: a kilogram of beef; a kilogram of bread; a smartphone; an automobile. Identify the dominant water-using process for each.
\end{exercise}

\begin{exercise}
Estimate the engineering workforce required to operate and maintain the water and wastewater infrastructure of a major city of 5 million people.
\end{exercise}

\subsection*{Simulation}\pathtag{standard}

\begin{exercise}
Implement a simulation of a multi-year drought on a reservoir-supplied city. Sweep the demand-management response stringency parameter and identify the conditions under which day-zero is avoided versus reached.
\end{exercise}

\begin{exercise}
Simulate the failure-cascade in a water distribution network under a major pressure-transient event. Identify the configurations that contain failures versus those that propagate them.
\end{exercise}

\begin{exercise}
Implement a climate-change scenario for a snowpack-dependent region. Sweep the warming parameter and identify the implications for reservoir operating rules.
\end{exercise}

\subsection*{Design}\pathtag{standard}

\begin{exercise}
Design a desalination plant for a city of $500{,}000$ people in a Mediterranean-climate coast. Specify capacity, technology choice (reverse osmosis vs thermal), energy source, brine disposal, integration with existing supply.
\end{exercise}

\begin{exercise}
Design a stormwater-management programme for a city facing increased precipitation intensity. Specify the green-infrastructure components (bioswales, permeable pavement, urban wetlands), the grey-infrastructure upgrades, the policy and zoning changes.
\end{exercise}

\begin{exercise}
Design a basin-scale water-allocation framework for a multi-jurisdiction river basin under climate-stressed conditions. Specify the allocation rules, the enforcement mechanism, the conflict-resolution procedures, the climate-adaptation provisions.
\end{exercise}

\subsection*{Judgement}\pathtag{mastery}

\begin{exercise}
Argue: the human right to water (UN General Assembly Resolution 64/292) imposes engineering and institutional obligations that the global water sector is not yet meeting. Argue the opposite. Identify the empirical evidence; identify the distributive and political considerations.
\end{exercise}

\begin{exercise}
Discuss the Flint water crisis as an engineering and institutional failure. Identify the specific engineering decisions, the institutional context, the distributive consequences. Identify what would have been required of the engineering profession to surface the consequences before they reached the public-health stage.
\end{exercise}

\begin{exercise}
Argue: desalination is the appropriate technology for water-stressed coastal cities; the energy cost and environmental footprint are acceptable in the climate-changed context. Argue the opposite. Identify the alternatives and the conditions under which each is preferable.
\end{exercise}

\begin{exercise}
Discuss the engineering profession's role in resolving multi-jurisdiction water disputes (Colorado River compact; Nile basin; Indus Waters Treaty under climate stress). Identify the technical contributions and the institutional limits.
\end{exercise}

\begin{exercise}
Discuss the Aral Sea collapse as a civilization-scale water-engineering failure. Identify the engineering decisions that produced it, the institutional context in which the consequences were not surfaced, and the implications for the engineering profession's responsibility when working under regimes that do not have channels for consequence escalation.
\end{exercise}

\begin{exercise}
Argue: agricultural-water efficiency programmes should be evaluated at basin scale rather than on-farm scale; on-farm efficiency improvements that reduce return flow can reduce downstream availability. Argue the opposite. Identify the implications for water-policy design.
\end{exercise}
